### 2012-03-18 ecoop12/rebuttal.txt
Initial paper move.

We thank the reviewers for their thorough critiques and thoughtful questions. We would like to particularly thank Reviewer 3 for soliciting an additional review from a researcher in the targeted user community.

Reviewer 1 expressed concern about the originality of Section 2. We agree that Section 2 does not introduce any individually novel design or adoption criteria. The purpose of this section was to bring together the issues important to language designers in this domain, which has never been done before. We do this to support our position that language designs should be evaluated holistically, against clear design and adoption criteria, consistent with the practices of other complex design disciplines. We will expand on this justification at the beginning of Section 2 in the final manuscript.

We acknowledge the concern raised by reviewers that Section 3 does not clearly state which aspects of the language design are individually novel. We believe that several of the language mechanisms we describe are in fact novel:

* The structural type inference scheme that cl.oquence uses has, to our knowledge, not been previously described. As described in the paper, within-function type inference is combined with type propagation between functions. Aspects of the within-function inference are controlled by the extension mechanism.
* To our knowledge, no language for GPU computing has provided kernel inference and no language targeting OpenCL has provided extension inference.
* The approach described for providing higher-order functions that are not first-class by inlining, rather than by the use of function pointers, has to our knowledge not been explored in existing languages (compilers sometimes do this, but languages do not guarantee this behavior.)
* Active libraries, although proposed previously, have not been adequately explored in practical languages (previous approaches relied on C++ template expansion). We expand on the concept of active libraries significantly, showing how a rich metalanguage (Python) enables a number of interesting language features (Sections 3.5-3.8).
* The language extension mechanism we describe is, to our knowledge, entirely novel, and is a major contribution of this paper.

We will take care to emphasize the extent to which each feature we describe is novel in the final manuscript.

Several reviewers note that we did not provide benchmarks to justify our claim that the OpenCL backend is performance-neutral. In fact, this claim is not an empirical one. We justify it by noting that all of the mechanisms that we describe are compile-time mechanisms -- cl.oquence does not introduce any additional run-time machinery or code transformations that may be the source of performance degradation. However, we would be happy to provide our benchmark data that demonstrates this 100% parity in performance to further emphasize this point in the final manuscript.

Several reviewers asked for more specific references to design criteria, in particular in section 3.2. We acknowledge this lack of specificity. Although we do reference section headings whenever relevant, we will be more specific about the way in which the described feature satisfies the referenced criteria in the final manuscript.

### 2012-03-18 csgf11/omar-vetter-csgf11.tex
Initial paper move.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% amspaper.tex --  LaTeX2e-based template for submissions to American 
% Meteorological Society Journals, including
%
% JAS 	-- Journal of the Atmospheric Sciences
% JAMC 	-- Journal of Applied Meteorology and Climatology
% JPO 	-- Journal of Physical Oceanography
% MWR 	-- Monthly Weather Review
% JTECH -- Journal of Atmospheric and Oceanic Technology
% WAF 	-- Weather and Forecasting
% JCLI 	-- Journal of Climate
% JHM 	-- Journal of Hydrometeorology
% JAM 	-- Journal of Applied Meteorology
%
% Template developed by B. Papa and S. Cooley, AMS. 
% Email questions to latex@ametsoc.org.
%
% August 12, 2008 (SRC)
%	- Clarified/added header notes, comments throughout
%	- Improved title page
%	- Edited text of document for clarity
%	- Altered list styles to adhere to AMS style, added comments
%	- Removed incorrect commands (i.e., \catcode) (corrects umlaut bug)
%	- Moved non-template commands to ametsoc.sty
%
% August, 2008 - B. Papa
% - Updated to handle two column journal page output
% - Updated text with new/modified instructions
%
% February, 2011 - B. Papa
% - Updated instructions for use with EM
%
% March, 2011 - B. Papa
% - Added use of \linenumbers
%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% PREAMBLE
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
% The following three commands will generate a PDF that follows all the requirements for submission
% and peer review.  Uncomment these commands to generate this output (and comment out the two lines below.)
%
% DOUBLE SPACE VERSION FOR SUBMISSION TO THE AMS
\documentclass[12pt]{article}
\usepackage{ametsoc}
%\linenumbers

%
% The following two commands will generate a single space, double column paper that closely
% matches an AMS journal page.  Uncomment these commands to generate this output (and comment
% out the two lines above. FOR AUTHOR USE ONLY. PAPERS SUBMITTED IN THIS FORMAT WILL BE RETURNED
% TO THE AUTHOR for submission with the correct formatting.
%
% TWO COLUMN JOURNAL PAGE LAYOUT FOR AUTHOR USE ONLY
%\documentclass[9pt]{article}
%\usepackage{ametsoc2col}

\input{macros}

%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% ABSTRACT
%
% Enter your Abstract here
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\newcommand{\myabstract}{From the perspective of computational scientists, programming systems are the most visible component of HPC systems and they serve a critical role in enabling HPC architectures that are high-performance, energy-efficient, scalable, robust, and productive. However, systems designed to achieve these goals must  handle increasing architectural complexity associated with hierarchical parallelism, the inertia associated with investments in legacy software and platforms and a variety of other design requirements. Many of these challenges are already emerging in today's multicore and heterogeneous computing systems and recent reports conclude that the road to Exascale Computing will require addressing these challenges even more directly. In this survey, we expand upon the design criteria that constrain programming systems for high-performance computing and highlight some emerging approaches, including recent research on general-purpose computing on graphics processors.}
%
\begin{document}
%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% TITLE
%
% Enter your TITLE here
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\title{\textbf{\large{Programming Systems on the Road to Exascale Computing
}}}
%
% Author names, with corresponding author information. 
% [Update and move the \thanks{...} block as appropriate.]
%
\author{\textsc{Cyrus Omar}\\
\textit{\footnotesize{Carnegie Mellon University}}
\and 
\centerline{\textsc{Jeffrey Vetter}}\\% Add additional authors, different insitution
\centerline{\textit{\footnotesize{Oak Ridge National Laboratory}}}
}
%
% The following block of code will handle the formatting of the title page depnding on whether
% we are formatting a double column (dc) author draft or a single column paper for submission.
% AUTHORS SHOULD SKIP OVER THIS... There is nothing to do in this section of code.
\ifthenelse{\boolean{dc}}
{
\twocolumn[
\begin{@twocolumnfalse}
\amstitle

% Start Abstract (Enter your Abstract above.  Do not enter any text here)
\begin{center}
\begin{minipage}{13.0cm}
\begin{abstract}
	\myabstract
	\newline
	\begin{center}
		\rule{38mm}{0.2mm}
	\end{center}
\end{abstract}
\end{minipage}
\end{center}
\end{@twocolumnfalse}
]
}
{
\amstitle
\begin{abstract}
\myabstract
\end{abstract}
\newpage
}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% MAIN BODY OF PAPER
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
\section{Introduction}
Computer-aided simulations and data analysis techniques have transformed science and engineering. Surveys show that scientists and engineers now spend more than 40\% of their time writing software \cite{howison2011scientific, hannay2009scientists}. Most of this software targets conventional desktop hardware, but about 20\% of scientists also target either local clusters or super\-computers for more numerically-intensive computations \cite{hannay2009scientists}, a number that continues to grow as scientists face ever-larger datasets and increasingly complex systems, such as those studied in the biological, environmental and social sciences.

Historically, scientists needing better performance simply purchased new hardware. This behavior was justified by appealing to Moore's Law, which has accurately predicted a doubling in the transistor count of integrated circuits every 18 months since the invention of the transistor in 1958 \cite{moore1998cramming}. Although Moore's Law remains valid, it is projected that by the end of this decade, the doubling period will increase substantially, due to fundamental physical limitations. More problematic, however, is that an assumed corollary of Moore's Law -- that {\it clock speeds} will also double every 18 months -- is no longer valid. Although transistor counts have continued to increase, clock speeds have remained flat for several years. It is now widely acknowledged that further performance gains will require that scientists make effective use of parallel computation.

Modern high-performance computing systems provide substantial hardware support for parallel computation. A single node in a cluster or a single workstation may contain several general-purpose multi-core processors as well as one or more  co-processors, such as a GPU, containing hundreds of specialized compute units. Modern clusters in turn  contain many such nodes -- the largest supercomputers today feature hundreds of thousands of nodes. These machines boast peak performance on the order of $10^{15}$ floating point operations per second (1 petaFLOPS). By the end of the decade, researchers hope to achieve a further thousand-fold increase in peak performance, a goal termed the {\it exascale initiative} \cite{exascale-roadmap}. At this scale, power consumption, hardware reliability, data movement and storage infrastructure become critical issues. Comprehensively addressing these issues  has emerged as one of the grand challenges in modern computing, and is a major focus of ongoing research spanning a variety of disciplines, including computer science, computer architecture, applied mathematics and a number of application domains in science and engineering.

In an ideal world, programmers would simply continue to write programs as they have in the past, relying on a compiler to optimize them for execution on a variety of hardware configurations at high speed. Unfortunately, such heroic compilers are likely impractical. Many problems central to automatic parallelization have been shown to be NP-Hard, or require additional information that can't be extracted from a program directly, so it is unlikely that fast, exact, fully-automatic techniques will emerge. Instead, optimization procedures must increasingly rely on heuristics, profiling data and domain-specific knowledge. Such so-called `fuzzy' approaches tend to require significant human involvement, meaning that programmers who hope to make full use of modern high-performance computing hardware must deal with complex issues beyond those they have faced in programming sequential hardware in the past. Recognition of this issue has led to calls to develop and improve the programming systems that developers rely on for assistance with these difficult issues. We use the term {\it programming systems} to encompass a broad range of tools
, including operating system APIs, programming languages, abstractions, libraries, compilers, editing environments, verification tools, documentation tools and so on.

The ultimate goal of this line of research is to maximize the end-to-end {\it productivity} of practitioners, measured by the time taken by all aspects of the process leading to a desired scientific result. In taking this view -- with its focus on human aspects of software development and design in addition to algorithmic and architectural considerations -- it is important to first identify and characterize the developer communities who will use these tools. Researchers typically classify software developers as either {\it end-user developers} or {\it professional developers}. End-user developers are those who have little formal training related to the tools that they use. These tools, as a consequence, tend to be relatively simple (spreadsheets, for example). Professional developers, on the other hand, have explicit training or extensive experience with software engineering practices and software development tools and techniques. Unfortunately, scientists and engineers do not cleanly fall into either of these groups. Although formal training in software development is rare \cite{oai:open.ac.uk.OAI2:17673}, scientists and engineers are more technically literate and demanding than typical end-users. For this reason, researchers now define a third group, {\it professional end-user developers} \cite{segal2007some}, to describe ``people working in highly technical, knowledge rich professions, such as financial mathematicians, scientists and engineers, who develop their own software in order to advance their own professional goals.''

In this survey, we examine the needs of professional end-user developers who currently work with, or will in the future need to work with, high-performance computing systems, including future exascale machines. As we do so, we describe both existing and emerging approaches, noting where additional research and development may be needed. We hope that the reader will emerge with a clear view of the state of the art and develop a clear, high-level understanding of the significant challenges remaining as we move toward future high-performance, high-productivity  computing systems and applications.

%\section{Programming Systems in HPC}
%Programmers today rely on a sophisticated collection of tools and abstractions to tackle the complex challenges faced in modern software engineering projects. In this section, we organize these programming systems and give a brief review of the choices most of these

\section{System Software}
Systems software is the portion of the programming system typically associated with the operating system or low-level runtime environment. It is responsible for interacting with hardware resources, usually in a relatively transparent way to the user. This category includes operating system APIs, file systems and I/O libraries, and systems management software \cite{exascale-roadmap}.

\subsection{Operating Systems} 
Today, some flavor of Linux is the dominant choice in HPC machines. Variants of UNIX, such as AIX, and lightweight kernel operating systems such as Compute Node Kernel (CNK) are also used by some vendors (notably, IBM and Cray) \cite{hpc-survey}. Operating system APIs are the most primitive abstractions available to programmers, so small choices (e.g. in thread scheduling) can have a large impact on performance. The choices made to accomodate conventional desktop and server workloads may not be ideal in an HPC context, so there has been considerable interest in developing HPC-specific operating systems. Exascale machines introduce additional complications, because resource management will require consideration of power budgets, and hardware reliability issues will become increasingly prominent. In some cases, collective decisions may need to be made based on information integrated across a cluster, further complicating matters.

Researchers differ about whether the simpler, more specialized lightweight kernel (LWK) operating systems dedicated to high-performance computing applications are better suited to exascale applications, outweighing the flexibility and broad developer base of systems based on a general-purpose operating system like Linux. In either case, however, there is a need for improved APIs for memory and thread management, performance monitoring and debugging, energy management and access to specialized hardware. To evaluate these APIs at scale before the associated hardware is available, whole-system simulation techniques must also be developed. These are all active research areas.

\subsection{File Systems and I/O Libraries}
Large, distributed file systems require specialized support at the systems level. A number of different high-performance file systems are in wide use today, including Lustre, GPFS and NFS. Together with I/O libraries such as netCDF, HDF5 and MPI-IO, these systems form the basis for data storage and retrieval on high-performance machines today \cite{hpc-survey}.

Unfortunately, it is not clear that these systems can continue to scale without serious changes. A major component of the overall energy budget of an exascale machine will be consumed by data movement. Today's systems are based on traditional file systems and continue to offer strong guarantees about conflicts, snchronization and coherence. I/O is also generally thought of as an activity that occurs periodically (e.g. during checkpointing) or at the end of a simulation, rather than as an activity that occurs on an ongoing basis throughout a computation. These constraining requirements and assumptions make scaling a challenge, even as new storage architectures and devices emerge, such as solid state disks. Fortunately, many applications do not require such strong guarantees in some cases. As more applications interleave data analysis with computation, relieving a major burden on I/O systems, they can be targeted to more specialized tasks such as checkpointing, improving their performance. As such, scalable, purpose-driven, hardware-aware I/O systems that are more tightly integrated into the programming language and environment are a major focus of ongoing research.

\subsection{Systems Management}
High-performance computing systems require additional management tools to ensure that needed resources are available and divided amongst users fairly and securely. For example, {\it batch systems} allow users to request jobs on shared machines, and schedule these jobs in an equitable way on available resources. Popular choices for batch systems  include Torque, MOAB, LoadLeveler and SLURM. A number of other tools are also used by systems administrators to maintain the cluster's software systems, monitor hardware and ensure long-term data integrity. Although there are fewer scaling challenges with this aspect of the software stack than with some of the others, these tools interact with several aspects of both the software stack and the hardware, and must be updated as these aspects of the high-performance computing systems change on the road to exascale.

\section{Programming Environments}
The programming environment consists of the software that developers interact with directly to develop and debug programs. This category centers around programming languages and includes parallel programming abstractions, domain-specific abstractions, runtime systems, compilers, debuggers, performance analysis tools, data analysis and visualization tools, and editors \cite{exascale-roadmap}.

Scientists and engineers generally prefer high-level scripting languages like MATLAB, Python, R and Perl \cite{nguyen2010survey}, particularly those bundled with powerful domain-specific libraries. These languages are typically interpreted rather than compiled and generally lack static type systems, relying instead on run-time (or ``dynamic'') type lookup and error handling. Although this may increase the flexibility of the language, these features also negatively impact performance and can lead to unexpected run-time errors that are generally impossible to rule out statically. This is particularly severe given that rigorous software testing practices are also quite rare in these domains \cite{nguyen2010survey}. When running at scale, a run-time error can be very costly, so better verification and testing of scientific code will be necessary.

Along code paths where performance appears to be a bottleneck, developers today turn to statically-typed, though still unsafe, low-level languages like C and Fortran \cite{4222616}. These languages give users explicit control over data layout and heap allocation and require explicit type annotations on all variables. Although this makes writing programs more tedious which can increase the number of logic errors a programmer makes, it can also potentially improve performance, particularly if the programmer understands the target architecture well (e.g. cache behavior). While a sequential reimplementation of a critical code path using a low-level language will generally produce a modest speedup, more significant speedups on modern hardware require parallelizing over many processor cores and, on massively-parallel machines, many interconnected nodes. Despite well-known difficulties, low-level, unsafe  approaches that use a form of shared-memory  multithreading (e.g. pthreads, OpenMP) paired with explicit message passing between nodes (typically using MPI) remain widely used for these tasks as well \cite{4222616, basili2008understanding}.

Although researchers often propose higher-level language features and new parallel programming abstractions that aim to strike a better balance between raw performance, productivity, code portability and verifiability (see below), end-users remain skeptical of new approaches. This viewpoint was perhaps most succinctly expressed by a participant interviewed in a recent study by Basili et al. \cite{basili2008understanding}, who stated ``I hate MPI, I hate C++. [But] if I had to choose again, I would probably choose the same.'' Although this sentiment is easy to dismiss as paradoxical, we believe that it demands direct examination by those in the research community working to advance the practice of scientific and high-performance computing. 
%
% Two types of languages see widespread use: productivity-oriented and performance-oriented
% Significant diversity in productivity-oriented languages. Valued for clean syntax and high-level abstractions. Usually very slow. Have a FFI to call into a performance-oriented language.
% Performance-oriented languages -- C or Fortran and their derivatives. Direct memory access, low-level abstractions, more verbose. Portable (though often need to include platform-specific optimizations, which can lead to messy code). Not easily extended with new abstractions and primitives (witness: lots of derivative languages that are not composable because no way to implement these abstractions cleanly)
% Generally don't support verifiability very well (performance-oriented languages have simple type systems); only exception is functional languages.
%

% Simplicity (Errors, Syntax, Core Semantics)
% Extensibility (New Abstractions, Domain-Specific Things, Domain- and Platform-Specific Optimizations)
% Performance (Need to allow control over data layout and movement, perhaps optionally)
% Portability (Architectures and operating systems; support new optimizations)
% Backwards-Compatibility (Existing codebases)
% Verifiability (type systems, metaprogramming-based, formal semantics, small TCB)
% Tool Support (syntax highlighting, editors, debuggers, etc.)
% Social Proof (Will not disappear - open source, significant projects have been developed, community)

As in many areas of design, it can be difficult to objectively evaluate the merit of language and tool designs. To better support such evaluations, researchers in design disciplines typically develop a set of high-level  design criteria that serve as a rubric for evaluating and guiding their design efforts. In programming language design, particularly for scientific and high-performance computing, there have been few concerted efforts to develop a coherent set of design criteria that capture the needs of the targeted developer communities. Similarly, there have been few treatments of adoption criteria for new languages and abstractions in practice, an important issue given the slow rates of adoption today. As such, we organize this section around a set of design and adoption criteria for new languages and abstractions based on prior empirical studies of this class of {\it professional end-user developer}, as well as our observations of characteristics common to successful projects in the past.

\subsection{Design Criteria}
We define design criteria as aspects of an abstraction or language's specification, rather than its implementation, that help developers express their intent naturally and correctly.

\subsubsection{Concise, Familiar and Readable Syntax}\label{syntax}
Despite a considerable amount of evidence pointing toward the value of a well-designed syntax, particularly for end-users in specialized domains \cite{pane1996usability}, the issue is sometimes marginalized. It is likely the case, however, that some libraries and languages experience low adoption due in  part due to the use of a verbose, unfamiliar or unreadable syntactic style.

A simple and concise syntax is common to most of the high-level languages that are widely used in scientific computing. Cordy identifies the principle of conciseness with elimination of redundancy and the availability of reasonable defaults \cite{cordy1992hints}. The high-level languages listed above have all either made optional, or removed entirely, much of the syntactic overhead characteristic of low-level languages, such as explicit variable declarations and extensive headers or preambles that contain information that can be inferred from the body of the program in most cases. They also typically feature simple array indexing syntax and concise literal forms for common data structures like arrays, matrices, sets and maps. A concise and minimal syntax eliminates unnecessary keystrokes and keeps more code visible on screen at a time. Studies have shown that there may be a correlation between lines of code entered and overall error rate, independent of other factors \cite{el2001confounding}.

However, the benefits of concise syntax must be balanced with concerns about readability and familiarity. A number of principles have been proposed to operationalize the notion of code readability. The most widely-used collection of principles are Green's cognitive dimensions \cite{green1996usability}. Of particular relevance is the notion of self-consistency, which serves to ensure that similar forms have similar meaning and that there are few subtle or context-dependent distinctions that users must be mindful of. Another important principle has been called {\it closeness of mapping}, expressing the value of a close correspondence between mental models and the formal model as expressed concretely using the language.

Most researchers become familiar with formal notation by studying mathematics. In nearly all commonly used languages in scientific computing, common mathematical notations or close approximations thereof are generally used. In contrast, many academic languages have settled on alternative notational styles. For example, the LISP family of languages uses a highly uniform list-based notation, while most functional languages typically borrow notation for function invocation from the lambda calculus, using the form \verb|f x| rather than \verb|f(x)| for function invocation. Although both of these styles have certain benefits, they can impose mental burdens on users who continue to mentally translate them into more familiar notation \cite{anderson1985novice}. Although these and related difficulties can decrease with experience, they remain an important barrier for new users to these languages.

Syntactic cues like whitespace and typography that do not have a formal meaning but rather exist to assist developers are called secondary notation \cite{green1990programming}. Most languages are whitespace-insensitive, while others (notably, Python) enforce consistent uses of whitespace, both in pursuit of consistency and as a technique to eliminate the need for block delimiters. Additionally, a few languages, such as Mathematica, have support for more typographically-rich mathematical notation via a structural editing interface. This technique appears to be helpful, although we do not know of formal studies that provide evidence for this claim.

\subsubsection{Support for Multiple Paradigms and Abstractions}\label{multiparadigm}
Although the difficulties of low-level and parallel programming are widely acknowledged, there is little consensus on which high-level abstractions are most appropriate for easing this burden. Indeed, it appears likely that no one abstraction will emerge triumphant over the others. Although library-based abstractions are useful, they often suffer from limitations of the language, particularly if they require compile-time support. Primitive language support for a parallel abstraction has been shown to be more useful than a library-based implementation in at least one case \cite{cave2010comparing}. As such, it is important that a language support several modes of operation, ideally without excessively favoring one over another.

Languages that have been designed specifically to explore a single abstraction as a core language feature often see limited adoption because they remain difficult to use in circumstances for which the favored abstraction is inappropriate. Examples of these abstractions and languages that feature them include:

\begin{itemize}
\item shared-memory concurrency (e.g. Java, C\#)
\item share-nothing message-passing (e.g. Erlang)
\item flat data and task parallelism (e.g. OpenMP and  OpenCL)
\item nested data and task parallelism (e.g. NESL, Copperhead)
\item transactional memory (e.g. the language described by Harris \cite{harris2003language})
\item automatic parallelization of functional primitives (e.g. Data Parallel Haskell, MapReduce and others)
\item (partitioned) global address spaces (e.g. UPC, Co-array Fortran, Global Arrays, Fortress, X10, Chapel and others)
\item adaptive thread migration (e.g. Charm++)
\end{itemize}

\subsubsection{Extensibility}\label{extensions}

Although support for multiple paradigms and primitive abstractions can be built into a language design, this leaves control in the hands of the language designer. Novel or domain-specific constructs that may be useful to a small number of users or in rare situations can be difficult to develop and distribute for this reason, leading to the proliferation of new languages as  described above. Language extensibility mechanisms support these use cases by giving users the ability to develop new abstractions and constructs that behave as if they were primitive constructs. If a mechanism is powerful enough, nearly all language constructs may be implemented using it, greatly simplifying the core semantics of a language.

Dynamic languages commonly rely on mechanisms like operator overloading and metaobject protocols \cite{Kiczales91} to provide extensibility via indirection. For example, the Python language allows objects to overload nearly every operator and as well as operations like attribute lookup (\verb|obj.attr|) and assignment. This mechanism is used by a number of libraries to create a more natural interface to a low-level API (e.g. \verb|pycuda|). More open-ended dynamic mechanisms, such as programmatic macros, have also been widely studied but as of yet have seen little adoption in languages used by professional end-users.

Language extensibility for statically-typed languages, on the other hand, remains an active research area. Compile-time metaprogramming systems, which are related to programmatic macros, have been developed for a number of languages (e.g. Template Haskell \cite{sheard2002template}) but these too have seen relatively limited development or adoption thus far.

Some researchers have advocated the use of compiler extension mechanisms, rather than mechanisms built into a language itself (cf. \cite{clements2008comparison}). Modern compilers now offer some support for front-end language extensions, although often these  can be quite difficult to use.  Although potentially powerful, this approach can also lead to issues when extensions are not easily composable or when multiple compilers exist for a language. Back-end extensions (that is, extensions that preserve the semantics of the language, improving only performance) are better supported in modern compilers.

Domain-specific language frameworks are a related approach that can serve many of the same goals as language-based  extension mechanisms \cite{fowler2010domain}. These tools ease the development of ``little languages'' and allow for the development and distribution of language features as modules. Domain experts use these tools to develop highly specialized languages that capture domain requirements precisely and allow for natural and concise specifications of scientific models and other structures. A major issue with this approach arises at language boundaries -- interoperability between different domain-specific languages is difficult due to feature mismatches. Another issue is that domain-specific languages often outgrow their initial implementations and begin to need increasingly powerful general-purpose features.

All extensibility mechanisms must be used carefully. Indeed, inexperienced developers can abuse mechanisms like operator overloading to create inscrutable interfaces that cause more problems than they solve. Some languages (notably Java) have taken up the philosophy that even simple language extension mechanisms should not be in the hands of end-users for this reason. Although this continues to be debated, we argue that extensibility is crucial for parallel and scientific programming in particular due to the significant levels of ongoing research into new parallel abstractions and the diverse set of domain-specific use cases.

\subsubsection{Support for Higher-Order Constructs}\label{hof}
A number of parallel programming data structures and algorithms are of higher-order, meaning that they take other functions or types as arguments or parameters. We argue that support for higher-order programming is critical to language usability in our target domains.

Well-known examples of functions that operate using other functions include fundamental parallel primitives like map, reduce and scan. Fortunately, most modern languages now support using functions as values. Languages like C and Fortran implement this using function pointers, although at considerable syntactic expense. However, unlike functional languages and most dynamic languages, they do not allow anonymous functions (that is, functions that are defined as inline expressions rather than statements), nor functions that close over variables in the surrounding scope. The most recent revision of C++ now has support for closures and closures are also being considered for a future revision of Java. OpenCL does not come with any support for higher-order functions, even via function pointers.

Functions and types that are parameterized by types are often referred to as {\it polymorphic} or {\it generic}. C++ supports this using its template system. Java, ML and Haskell support a simpler form of {\it parametric polymorphism}, and Haskell also supports a more flexible {\it type class} system. C++, Java and Fortran also support {\it function overloading}, allowing multiple versions of a function that differ only according to their argument types. C and OpenCL do not support any form of polymorphism or user-defined function overloading.

Dynamic languages do not require that variables be assigned a type, so generic functions are written using run-time checks that ensure that a particular value supports the specific interface that a function expects. This is sometimes known as ``duck typing'' and is generally considered as a useful feature, due to its considerable flexibility. A promising approach that resembles duck typing, while operating statically, is known as {\it structural typing} \cite{malayeri2009structural}.

\subsubsection{Modularity and Packaging}\label{modularity}
Modularity is a broad term that refers to mechanisms useful for combining and reusing independently developed libraries of code. Languages with good support for modularity promote information hiding, thus localizing the effect of code changes, and allow modules to communicate over well-defined interfaces. There remains widespread disagreement and considerable ongoing research on language support for modularity. Object-oriented methodologies, although common in industrial projects, are used less frequently in scientific codes due to a perceived loss of control and performance \cite{basili2008understanding}. Many dynamic languages support modularity only implicitly using duck typing. Functional languages, on the other hand, often have module systems that enforce correct usage of an interface at compile-time, albeit at some notational cost \cite{tapl}.

The packaging and linking mechanisms available in a language can also significantly impact its usability. Languages like C and C++ use a fragile preprocessor-based packaging mechanism and require separate header files, which often leads to subtle errors and compilation inefficiencies. More recent high-level languages have developed a varied set of mechanisms that are significantly simpler for packaging and deployment.

\subsubsection{Verifiability}\label{verifiability}
Verifying that a program does not contain errors and that it will operate according to specification (if a specification exists, which can be rare in this domain \cite{oai:open.ac.uk.OAI2:17673}) is a critical concern across all areas of software development. Errors in scientific programs may arise due to problems in basic program logic, as in other domains, but also due to the accumulation of  numerical approximation errors or by violation of domain-specific constraints (e.g. inconsistent scientific units.)

A number of design decisions can influence the difficulty of formal program verification. Broadly stated, unconstrained support for dynamic indirection and direct access to memory have made program verification more difficult in many commonly-used languages. As a result, many classes of errors are only caught at run-time, often due to edge cases that are difficult to test for.

Advanced type systems, matched with appropriately constrained data structures and control constructs, can dramatically increase the likelihood that an error is found at compile-time and even eliminate entire classes of errors \cite{tapl}. A large portion of academic research on programming language design has focused on designing such type systems. Unfortunately, as with many new parallel abstractions, these generally require that a language be modified with new primitives and most work has either been with prototype languages or functional languages like Ocaml, Haskell or Coq, which, due to some of the factors mentioned here, have seen  limited (though growing) adoption in scientific computing so far.

In the absence of a formal proof of correctness, users must test programs with specific inputs, relying both on language-provided run-time checks and user-provided run-time assertions to catch errors. Most languages have unit testing frameworks designed to minimize the burden of developing unit tests, though these are used infrequently in science \cite{oai:open.ac.uk.OAI2:17673, hannay2009scientists}. Several languages (e.g. Eiffel) also have support for specifying pre- and post-conditions for functions, so that the assertions that must hold about arguments are visible in the function signature, but this feature has not been well-supported by widely-used languages to date.

\subsection{Adoption Criteria}
While most of the design criteria described in the previous section have been the topic of significant (and ongoing) research, there are a number of other, more practical issues that can significantly affect adoption of a new language or tool. It should be noted that certain design decisions may make it easier to satisfy these adoption to criteria as well, an important consideration given that incentive structures in academia often do not reward efforts to improve a language along some of the following dimensions \cite{howison2011scientific}.

\subsubsection{Performance}
Performance is, of course, a critical concern in scientific and (particularly \cite{basili2008understanding}) high-performance computing. Low-level languages typically elect to give the user direct access to hardware primitives. High-level languages must rely on a sophisticated compiler to translate high-level abstractions into performant code. Although the latter approach would be ideal, program optimization remains an active research area with many open problems. Indeed, many relevant algorithms have been shown to be NP-Complete, so compilers must rely on heuristics. Humans remain better at inventing and applying heuristics, given enough motivation and experience, than computers in many cases. Indeed, even when the compiler can sometimes produce better optimizations, users generally insist on being able to form a mental model of what their code is doing at the machine-level so that they can {\it reason} about the effects of code changes on overall performance \cite{squires2005programmers}. As such, users generally demand that low-level abstractions remain available alongside high-level abstractions. It may also be useful to support a model where programmers can formalize, package and directly apply optimization heuristics programmatically, rather than relying on a ``black-box'' compiler.

% Squires et al, 2005 -- programmers want to know what is happening under the hood so they can reason about optimization

\subsubsection{Portability}\label{portability}
A number of surveys have revealed that portability is one of the most important issues considered by scientists and engineers \cite{hannay2009scientists}\cite{basili2008understanding}. Several processor architectures and operating systems are in widespread use, with more appearing on a fairly regular basis. Low-level languages like C have generally achieved a reasonable level of portability, although aspects of the language that are underspecified can be a source of errors. The OpenCL language was designed with portability as a major design criteria, and compilers for OpenCL are now available for a diverse collection of processor and accelerator architectures. High-level languages are generally highly portable due to the use of a virtual machine architecture.

It should be noted, however, that simple portability is not entirely sufficient -- users want to be able to write programs that can be ported to new architectures and operating systems and achieve high performance without significant retuning. This remains an area of active research. Recent efforts to build abstractions that generate efficient code for devices with multi-level memories, as implemented in the Sequoia language \cite{fatahalian2006sequoia}, have  progressed toward this goal.

\subsubsection{Useful Error Messages}\label{errors}
When the compiler or run-time system of a language generates an error, users must determine the source of the problem using the information provided in an error message. Studies have shown that good error messages can dramatically reduce the time it takes to debug programs \cite{marceau2011measuring}. Indeed, a widely-reported frustration with C++ is that it produces overly verbose and cryptic error messages, particularly when using templates.

\subsubsection{Tool and Infrastructure Support}\label{tools}
% Squires 2005 might have a ref
Modern software development now relies on a number of tools to assist with common tasks, such as debuggers,  profilers, syntax-aware editors, documentation generators, style checkers and interactive interpreters. Similarly, most established languages benefit from a centralized package repository (e.g. PyPI) paired with a package manager that can automate the installation of new packages. These tools and infrastructure are not trivial to develop from scratch, and few incentives exist for researchers to do so, but few users are willing to do without these tools for non-trivial projects \cite{squires2005programmers}.

\subsubsection{Backwards Compatibility}
End-user communities have produced a number of highly tuned and tested packages that are widely used in their fields. Porting these libraries to a new language requires significant effort and few communities will be willing to do so, particularly before a language is very well established. There is generally a greater willingness to develop wrappers that invoke functions from these packages, however. A powerful foreign function interface, ideally able to handle native code as well as code in existing widely-used high-level languages, enables these efforts to proceed smoothly and can provide a language with a large package library relatively early in its development.

Many of the reasonably successful approaches to date are built incrementally on existing languages as libraries or as simple  extensions to existing languages (e.g. OpenMP, UPC, Co-array Fortran, Cilk++, MPI, CUDA, OpenCL). Notably, this allows existing programs to continue to function correctly, while enabling the gradual integration of the new constructs in parts of the code that would derive the greatest benefit. In contrast, approaches that have required total rewrites have had more difficulty recruiting early adopters (e.g. X10 \cite{charles2005x10}.)

\subsubsection{Learning Material}
Existing languages benefit from a large collection of books, presentations and tutorials that allow new users to get started quickly. The most mature languages benefit further from the availability of courses and professional training seminars. New languages often suffer due to the lack of such polished learning material, or in some cases, due to the lack of {\it any} significant learning material targeted at end users (rather than other language designers.)

\subsubsection{Open Availability}
New languages and abstractions are often viewed with suspicion if they do not come with source code that can be modified under a free software license. The most unencumbered licenses (other than public domain releases) are BSD-style licenses. So called {\it copyleft} licenses like the GPL are also popular, requiring primarily that modifications to the compiler be distributed under the same license.

\subsubsection{Social Proof}
Finally, users look to the language's user community for evidence that the language is a serious effort that will not be abandoned and that there are other developers building libraries and sharing information over established communication channels. The availability of non-trivial demonstrations have been cited in surveys as critical to adoption as well \cite{basili2008understanding}. These criteria, as with many of the others described above, can be difficult to satisfy, particularly for ``clean-slate'' language designs, which may help to explain why language adoption is often slow. However, even projects with a narrower scope, such as new parallel or domain-specific abstractions, must tackle these issues. Funding agencies can aid in adoption by guaranteeing development will be supported for a number of years for languages that appear to have the greatest promise.

% What Do Programmers of Parallel Machines Need?
% Squiers et al, 2005
% - Global view
% - Tool support
% - Understanding of what the code is doing
% 
%As should be clear, there are many challenges facing any new language design effort. Many of these challenges apply also to efforts with a narrower scope (e.g. a new parallel or domain-specific abstraction). Although the research community continues to tackle open issues described as design criteria, the challenges described as adoption criteria above have not generally received as much attention.

% An example of a floating figure using the graphicx package.
% Note that \label must occur AFTER (or within) \caption.
% For figures, \caption should occur after the \includegraphics.
% Note that IEEEtran v1.7 and later has special internal code that
% is designed to preserve the operation of \label within \caption
% even when the captionsoff option is in effect. However, because
% of issues like this, it may be the safest practice to put all your
% \label just after \caption rather than within \caption{}.
%
% Reminder: the "draftcls" or "draftclsnofoot", not "draft", class
% option should be used if it is desired that the figures are to be
% displayed while in draft mode.
%
%\begin{figure}[!t]
%\centering
%\includegraphics[width=2.5in]{myfigure}
% where an .eps filename suffix will be assumed under latex, 
% and a .pdf suffix will be assumed for pdflatex; or what has been declared
% via \DeclareGraphicsExtensions.
%\caption{Simulation Results}
%\label{fig_sim}
%\end{figure}

% Note that IEEE typically puts floats only at the top, even when this
% results in a large percentage of a column being occupied by floats.

% An example of a double column floating figure using two subfigures.
% (The subfig.sty package must be loaded for this to work.)
% The subfigure \label commands are set within each subfloat command, the
% \label for the overall figure must come after \caption.
% \hfil must be used as a separator to get equal spacing.
% The subfigure.sty package works much the same way, except \subfigure is
% used instead of \subfloat.
%
%\begin{figure*}[!t]
%\centerline{\subfloat[Case I]\includegraphics[width=2.5in]{subfigcase1}%
%\label{fig_first_case}}
%\hfil
%\subfloat[Case II]{\includegraphics[width=2.5in]{subfigcase2}%
%\label{fig_second_case}}}
%\caption{Simulation results}
%\label{fig_sim}
%\end{figure*}
%
% Note that often IEEE papers with subfigures do not employ subfigure
% captions (using the optional argument to \subfloat), but instead will
% reference/describe all of them (a), (b), etc., within the main caption.

% An example of a floating table. Note that, for IEEE style tables, the 
% \caption command should come BEFORE the table. Table text will default to
% \footnotesize as IEEE normally uses this smaller font for tables.
% The \label must come after \caption as always.
%
%\begin{table}[!t]
%% increase table row spacing, adjust to taste
%\renewcommand{\arraystretch}{1.3}
% if using array.sty, it might be a good idea to tweak the value of
% \extrarowheight as needed to properly center the text within the cells
%\caption{An Example of a Table}
%\label{table_example}
%\centering
%% Some packages, such as MDW tools, offer better commands for making tables
%% than the plain LaTeX2e tabular which is used here.
%\begin{tabular}{|c||c|}
%\hline
%One & Two\\
%\hline
%Three & Four\\
%\hline
%\end{tabular}
%\end{table}

% Note that IEEE does not put floats in the very first column - or typically
% anywhere on the first page for that matter. Also, in-text middle ("here")
% positioning is not used. Most IEEE journals/conferences use top floats
% exclusively. Note that, LaTeX2e, unlike IEEE journals/conferences, places
% footnotes above bottom floats. This can be corrected via the \fnbelowfloat
% command of the stfloats package.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% REFERENCES
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Create a bibliography directory and place your .bib file there.
\ifthenelse{\boolean{dc}}
{}
{\clearpage}
%\bibliographystyle{ametsoc}
%\bibliographystyle{IEEEtran}
\bibliographystyle{plain-annote}
\bibliography{../research}

\end{document}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% END OF TEMPLATE
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

### 2012-03-18 icse12/graphite-icse12.tex
Initial paper move.

%% bare_conf.tex
%% V1.3
%% 2007/01/11
%% by Michael Shell
%% See:
%% http://www.michaelshell.org/
%% for current contact information.
%%
%% This is a skeleton file demonstrating the use of IEEEtran.cls
%% (requires IEEEtran.cls version 1.7 or later) with an IEEE conference paper.
%%
%% Support sites:
%% http://www.michaelshell.org/tex/ieeetran/
%% http://www.ctan.org/tex-archive/macros/latex/contrib/IEEEtran/
%% and
%% http://www.ieee.org/

%%*************************************************************************
%% Legal Notice:
%% This code is offered as-is without any warranty either expressed or
%% implied; without even the implied warranty of MERCHANTABILITY or
%% FITNESS FOR A PARTICULAR PURPOSE! 
%% User assumes all risk.
%% In no event shall IEEE or any contributor to this code be liable for
%% any damages or losses, including, but not limited to, incidental,
%% consequential, or any other damages, resulting from the use or misuse
%% of any information contained here.
%%
%% All comments are the opinions of their respective authors and are not
%% necessarily endorsed by the IEEE.
%%
%% This work is distributed under the LaTeX Project Public License (LPPL)
%% ( http://www.latex-project.org/ ) version 1.3, and may be freely used,
%% distributed and modified. A copy of the LPPL, version 1.3, is included
%% in the base LaTeX documentation of all distributions of LaTeX released
%% 2003/12/01 or later.
%% Retain all contribution notices and credits.
%% ** Modified files should be clearly indicated as such, including  **
%% ** renaming them and changing author support contact information. **
%%
%% File list of work: IEEEtran.cls, IEEEtran_HOWTO.pdf, bare_adv.tex,
%%                    bare_conf.tex, bare_jrnl.tex, bare_jrnl_compsoc.tex
%%*************************************************************************

% *** Authors should verify (and, if needed, correct) their LaTeX system  ***
% *** with the testflow diagnostic prior to trusting their LaTeX platform ***
% *** with production work. IEEE's font choices can trigger bugs that do  ***
% *** not appear when using other class files.                            ***
% The testflow support page is at:
% http://www.michaelshell.org/tex/testflow/

% Note that the a4paper option is mainly intended so that authors in
% countries using A4 can easily print to A4 and see how their papers will
% look in print - the typesetting of the document will not typically be
% affected with changes in paper size (but the bottom and side margins will).
% Use the testflow package mentioned above to verify correct handling of
% both paper sizes by the user's LaTeX system.
%
% Also note that the "draftcls" or "draftclsnofoot", not "draft", option
% should be used if it is desired that the figures are to be displayed in
% draft mode.
%
\documentclass[10pt, conference, compsocconf]{IEEEtran}
% Add the compsocconf option for Computer Society conferences.
%
% If IEEEtran.cls has not been installed into the LaTeX system files,
% manually specify the path to it like:
% \documentclass[conference]{../sty/IEEEtran}

\usepackage{rotating}

% Some very useful LaTeX packages include:
% (uncomment the ones you want to load)

% *** MISC UTILITY PACKAGES ***
%
%\usepackage{ifpdf}
% Heiko Oberdiek's ifpdf.sty is very useful if you need conditional
% compilation based on whether the output is pdf or dvi.
% usage:
% \ifpdf
%   % pdf code
% \else
%   % dvi code
% \fi
% The latest version of ifpdf.sty can be obtained from:
% http://www.ctan.org/tex-archive/macros/latex/contrib/oberdiek/
% Also, note that IEEEtran.cls V1.7 and later provides a builtin
% \ifCLASSINFOpdf conditional that works the same way.
% When switching from latex to pdflatex and vice-versa, the compiler may
% have to be run twice to clear warning/error messages.

% *** CITATION PACKAGES ***
%
%\usepackage{cite}
% cite.sty was written by Donald Arseneau
% V1.6 and later of IEEEtran pre-defines the format of the cite.sty package
% \cite{} output to follow that of IEEE. Loading the cite package will
% result in citation numbers being automatically sorted and properly
% "compressed/ranged". e.g., [1], [9], [2], [7], [5], [6] without using
% cite.sty will become [1], [2], [5]--[7], [9] using cite.sty. cite.sty's
% \cite will automatically add leading space, if needed. Use cite.sty's
% noadjust option (cite.sty V3.8 and later) if you want to turn this off.
% cite.sty is already installed on most LaTeX systems. Be sure and use
% version 4.0 (2003-05-27) and later if using hyperref.sty. cite.sty does
% not currently provide for hyperlinked citations.
% The latest version can be obtained at:
% http://www.ctan.org/tex-archive/macros/latex/contrib/cite/
% The documentation is contained in the cite.sty file itself.

% *** GRAPHICS RELATED PACKAGES ***
%
\ifCLASSINFOpdf
  % \usepackage[pdftex]{graphicx}
  % declare the path(s) where your graphic files are
  % \graphicspath{{../pdf/}{../jpeg/}}
  % and their extensions so you won't have to specify these with
  % every instance of \includegraphics
  % \DeclareGraphicsExtensions{.pdf,.jpeg,.png}
\else
  % or other class option (dvipsone, dvipdf, if not using dvips). graphicx
  % will default to the driver specified in the system graphics.cfg if no
  % driver is specified.
  % \usepackage[dvips]{graphicx}
  % declare the path(s) where your graphic files are
  % \graphicspath{{../eps/}}
  % and their extensions so you won't have to specify these with
  % every instance of \includegraphics
  % \DeclareGraphicsExtensions{.eps}
\fi
% graphicx was written by David Carlisle and Sebastian Rahtz. It is
% required if you want graphics, photos, etc. graphicx.sty is already
% installed on most LaTeX systems. The latest version and documentation can
% be obtained at: 
% http://www.ctan.org/tex-archive/macros/latex/required/graphics/
% Another good source of documentation is "Using Imported Graphics in
% LaTeX2e" by Keith Reckdahl which can be found as epslatex.ps or
% epslatex.pdf at: http://www.ctan.org/tex-archive/info/
%
% latex, and pdflatex in dvi mode, support graphics in encapsulated
% postscript (.eps) format. pdflatex in pdf mode supports graphics
% in .pdf, .jpeg, .png and .mps (metapost) formats. Users should ensure
% that all non-photo figures use a vector format (.eps, .pdf, .mps) and
% not a bitmapped formats (.jpeg, .png). IEEE frowns on bitmapped formats
% which can result in "jaggedy"/blurry rendering of lines and letters as
% well as large increases in file sizes.
%
% You can find documentation about the pdfTeX application at:
% http://www.tug.org/applications/pdftex

% *** MATH PACKAGES ***
%
%\usepackage[cmex10]{amsmath}
% A popular package from the American Mathematical Society that provides
% many useful and powerful commands for dealing with mathematics. If using
% it, be sure to load this package with the cmex10 option to ensure that
% only type 1 fonts will utilized at all point sizes. Without this option,
% it is possible that some math symbols, particularly those within
% footnotes, will be rendered in bitmap form which will result in a
% document that can not be IEEE Xplore compliant!
%
% Also, note that the amsmath package sets \interdisplaylinepenalty to 10000
% thus preventing page breaks from occurring within multiline equations. Use:
%\interdisplaylinepenalty=2500
% after loading amsmath to restore such page breaks as IEEEtran.cls normally
% does. amsmath.sty is already installed on most LaTeX systems. The latest
% version and documentation can be obtained at:
% http://www.ctan.org/tex-archive/macros/latex/required/amslatex/math/

% *** SPECIALIZED LIST PACKAGES ***
%
%\usepackage{algorithmic}
% algorithmic.sty was written by Peter Williams and Rogerio Brito.
% This package provides an algorithmic environment fo describing algorithms.
% You can use the algorithmic environment in-text or within a figure
% environment to provide for a floating algorithm. Do NOT use the algorithm
% floating environment provided by algorithm.sty (by the same authors) or
% algorithm2e.sty (by Christophe Fiorio) as IEEE does not use dedicated
% algorithm float types and packages that provide these will not provide
% correct IEEE style captions. The latest version and documentation of
% algorithmic.sty can be obtained at:
% http://www.ctan.org/tex-archive/macros/latex/contrib/algorithms/
% There is also a support site at:
% http://algorithms.berlios.de/index.html
% Also of interest may be the (relatively newer and more customizable)
% algorithmicx.sty package by Szasz Janos:
% http://www.ctan.org/tex-archive/macros/latex/contrib/algorithmicx/

% *** ALIGNMENT PACKAGES ***
%
%\usepackage{array}
% Frank Mittelbach's and David Carlisle's array.sty patches and improves
% the standard LaTeX2e array and tabular environments to provide better
% appearance and additional user controls. As the default LaTeX2e table
% generation code is lacking to the point of almost being broken with
% respect to the quality of the end results, all users are strongly
% advised to use an enhanced (at the very least that provided by array.sty)
% set of table tools. array.sty is already installed on most systems. The
% latest version and documentation can be obtained at:
% http://www.ctan.org/tex-archive/macros/latex/required/tools/

%\usepackage{mdwmath}
%\usepackage{mdwtab}
% Also highly recommended is Mark Wooding's extremely powerful MDW tools,
% especially mdwmath.sty and mdwtab.sty which are used to format equations
% and tables, respectively. The MDWtools set is already installed on most
% LaTeX systems. The lastest version and documentation is available at:
% http://www.ctan.org/tex-archive/macros/latex/contrib/mdwtools/

% IEEEtran contains the IEEEeqnarray family of commands that can be used to
% generate multiline equations as well as matrices, tables, etc., of high
% quality.

%\usepackage{eqparbox}
% Also of notable interest is Scott Pakin's eqparbox package for creating
% (automatically sized) equal width boxes - aka "natural width parboxes".
% Available at:
% http://www.ctan.org/tex-archive/macros/latex/contrib/eqparbox/

% *** SUBFIGURE PACKAGES ***
%\usepackage[tight,footnotesize]{subfigure}
% subfigure.sty was written by Steven Douglas Cochran. This package makes it
% easy to put subfigures in your figures. e.g., "Figure 1a and 1b". For IEEE
% work, it is a good idea to load it with the tight package option to reduce
% the amount of white space around the subfigures. subfigure.sty is already
% installed on most LaTeX systems. The latest version and documentation can
% be obtained at:
% http://www.ctan.org/tex-archive/obsolete/macros/latex/contrib/subfigure/
% subfigure.sty has been superceeded by subfig.sty.

%\usepackage[caption=false]{caption}
%\usepackage[font=footnotesize]{subfig}
% subfig.sty, also written by Steven Douglas Cochran, is the modern
% replacement for subfigure.sty. However, subfig.sty requires and
% automatically loads Axel Sommerfeldt's caption.sty which will override
% IEEEtran.cls handling of captions and this will result in nonIEEE style
% figure/table captions. To prevent this problem, be sure and preload
% caption.sty with its "caption=false" package option. This is will preserve
% IEEEtran.cls handing of captions. Version 1.3 (2005/06/28) and later 
% (recommended due to many improvements over 1.2) of subfig.sty supports
% the caption=false option directly:
%\usepackage[caption=false,font=footnotesize]{subfig}
%
% The latest version and documentation can be obtained at:
% http://www.ctan.org/tex-archive/macros/latex/contrib/subfig/
% The latest version and documentation of caption.sty can be obtained at:
% http://www.ctan.org/tex-archive/macros/latex/contrib/caption/

% *** FLOAT PACKAGES ***
%
%\usepackage{fixltx2e}
% fixltx2e, the successor to the earlier fix2col.sty, was written by
% Frank Mittelbach and David Carlisle. This package corrects a few problems
% in the LaTeX2e kernel, the most notable of which is that in current
% LaTeX2e releases, the ordering of single and double column floats is not
% guaranteed to be preserved. Thus, an unpatched LaTeX2e can allow a
% single column figure to be placed prior to an earlier double column
% figure. The latest version and documentation can be found at:
% http://www.ctan.org/tex-archive/macros/latex/base/

%\usepackage{stfloats}
% stfloats.sty was written by Sigitas Tolusis. This package gives LaTeX2e
% the ability to do double column floats at the bottom of the page as well
% as the top. (e.g., "\begin{figure*}[!b]" is not normally possible in
% LaTeX2e). It also provides a command:
%\fnbelowfloat
% to enable the placement of footnotes below bottom floats (the standard
% LaTeX2e kernel puts them above bottom floats). This is an invasive package
% which rewrites many portions of the LaTeX2e float routines. It may not work
% with other packages that modify the LaTeX2e float routines. The latest
% version and documentation can be obtained at:
% http://www.ctan.org/tex-archive/macros/latex/contrib/sttools/
% Documentation is contained in the stfloats.sty comments as well as in the
% presfull.pdf file. Do not use the stfloats baselinefloat ability as IEEE
% does not allow \baselineskip to stretch. Authors submitting work to the
% IEEE should note that IEEE rarely uses double column equations and
% that authors should try to avoid such use. Do not be tempted to use the
% cuted.sty or midfloat.sty packages (also by Sigitas Tolusis) as IEEE does
% not format its papers in such ways.

% *** PDF, URL AND HYPERLINK PACKAGES ***
%
%\usepackage{url}
% url.sty was written by Donald Arseneau. It provides better support for
% handling and breaking URLs. url.sty is already installed on most LaTeX
% systems. The latest version can be obtained at:
% http://www.ctan.org/tex-archive/macros/latex/contrib/misc/
% Read the url.sty source comments for usage information. Basically,
% \url{my_url_here}.

% *** Do not adjust lengths that control margins, column widths, etc. ***
% *** Do not use packages that alter fonts (such as pslatex).         ***
% There should be no need to do such things with IEEEtran.cls V1.6 and later.
% (Unless specifically asked to do so by the journal or conference you plan
% to submit to, of course. )

% correct bad hyphenation here
\hyphenation{op-tical net-works semi-conduc-tor}

\usepackage{alltt}
\usepackage{hyperref}
\hypersetup{pdfborder={0 0 0 0}}

\begin{document}
%
% paper title
% can use linebreaks \\ within to get better formatting as desired
\title{Active Code Completion}

% author names and affiliations
% use a multiple column layout for up to two different
% affiliations

\author{\IEEEauthorblockN{Cyrus Omar, YoungSeok Yoon, Thomas D. LaToza, Brad A. Myers}
\IEEEauthorblockA{Carnegie Mellon University, 
Pittsburgh, PA, USA\\
\small{\texttt{\{comar,youngseok,tlatoza,bam\}$@$cs.cmu.edu}}\\
\small{\texttt{\url{http://www.cs.cmu.edu/~NatProg/graphite.html}}}}
}

% conference papers do not typically use \thanks and this command
% is locked out in conference mode. If really needed, such as for
% the acknowledgment of grants, issue a \IEEEoverridecommandlockouts
% after \documentclass

% for over three affiliations, or if they all won't fit within the width
% of the page, use this alternative format:
% 
%\author{\IEEEauthorblockN{Michael Shell\IEEEauthorrefmark{1},
%Homer Simpson\IEEEauthorrefmark{2},
%James Kirk\IEEEauthorrefmark{3}, 
%Montgomery Scott\IEEEauthorrefmark{3} and
%Eldon Tyrell\IEEEauthorrefmark{4}}
%\IEEEauthorblockA{\IEEEauthorrefmark{1}School of Electrical and Computer Engineering\\
%Georgia Institute of Technology,
%Atlanta, Georgia 30332--0250\\ Email: see http://www.michaelshell.org/contact.html}
%\IEEEauthorblockA{\IEEEauthorrefmark{2}Twentieth Century Fox, Springfield, USA\\
%Email: homer@thesimpsons.com}
%\IEEEauthorblockA{\IEEEauthorrefmark{3}Starfleet Academy, San Francisco, California 96678-2391\\
%Telephone: (800) 555--1212, Fax: (888) 555--1212}
%\IEEEauthorblockA{\IEEEauthorrefmark{4}Tyrell Inc., 123 Replicant Street, Los Angeles, California 90210--4321}}

% use for special paper notices
%\IEEEspecialpapernotice{(Invited Paper)}

% make the title area
\maketitle

\begin{abstract}
Code completion menus have replaced standalone API browsers for most developers because they are more tightly  integrated into the development workflow. Refinements to the code completion menu that incorporate additional sources of information have similarly been shown to be valuable, even relative to standalone counterparts offering similar functionality. In this paper, we describe {\em active code completion}, a technique that enables the integration of a much broader array of interactive and highly-specialized code generation interfaces, called {\em palettes}, directly into the editor environment. Using several empirical methods, we examine the contexts in which such a system could be useful, describe the design constraints governing the system architecture as well as particular code completion interfaces, and design one such system, named Graphite, for the Eclipse Java development environment. Using Graphite, we implement a palette for writing regular expressions as our primary  example and conduct a controlled user study that provides further evidence in support of the claim that integrating specialized tools directly into the editor environment is both feasible, particularly if aided by an  active code completion system like Graphite, and valuable to professional developers.
\end{abstract}

\begin{IEEEkeywords}
code completion, development environments

\end{IEEEkeywords}

% For peer review papers, you can put extra information on the cover
% page as needed:
% \ifCLASSOPTIONpeerreview
% \begin{center} \bfseries EDICS Category: 3-BBND \end{center}
% \fi
%
% For peerreview papers, this IEEEtran command inserts a page break and
% creates the second title. It will be ignored for other modes.
\IEEEpeerreviewmaketitle
\section{Introduction}
Software developers today make heavy use of the code completion support found in modern source code editors  \cite{murphy_how_2006}. Most editors provide code completion in the form of a floating menu containing  contextually-relevant variables, fields, methods, types and other code snippets. By navigating and selecting from this menu, developers are able to avoid many common spelling and logic errors, eliminate unnecessary keystrokes and explore unfamiliar APIs without incurring the mental overhead associated with switching to an external documentation tool or API browser.
% The way code completion menus brought documentation systems into the editor, active code completion interfaces bring external tools into the editor.

Several refinements and additions to the code completion menu have previously been suggested in the literature. These have focused on leveraging additional sources of information, such as databases of usage history \cite{robbes_how_2008}, examples extracted from code repositories \cite{bruch_learning_2009} and crowdsourced information \cite{mooty_calcite:_2010}, to increase the relevance and sophistication of the featured menu items. As with the standard form of code completion, many of these sources of data can also be utilized via external tools (e.g. Calcite \cite{mooty_calcite:_2010} uses information that could already be accessed using the Jadeite \cite{conf/vl/StylosFYM09} tool). Empirical evidence presented in these studies, however, indicates that directly integrating these kinds of tools directly into the editor environment can be more effective.

In all of these systems, the code completion interface has remained primarily menu-based. When an item is selected, code is inserted immediately, without further input from the developer. These systems are also limited in that they are difficult to modularly extend. A fixed strategy determines the completions that are available, leaving library providers unable to specify new domain-specific or contextually-relevant items. 
In this paper we propose a technique called {\it active code completion} that eliminates these restrictions. This  makes developing and integrating a broad array of highly-specialized developer tools directly into the editor, via the familiar code completion command, significantly simpler. This technique is motivated by the evidence discussed above and further evidence provided in this paper that developers prefer, and make more effective use of, tools that do not require leaving the immediate editing environment.

In this paper, we discuss active code completion in the context of object construction. For example, consider the following Java code fragment:

\begin{alltt}
  public Color getDefaultColor() \{
      return \textvisiblespace
\end{alltt}

\begin{figure*}\label{color}
\begin{center}
\includegraphics[scale=0.8]{color_palette.png}\end{center}
\caption{(a) An example code completion palette associated with the \texttt{Color} class. (b) The source code generated by this palette.}
\end{figure*}

If the developer invokes the code completion command at the indicated cursor position, the editor  looks  for a {\it palette definition} associated with the {\it type} of the expression being entered, which in this case is  \verb|Color|. If an associated palette is found, a menu item briefly describing this palette is added to the standard code completion menu. When selected, the corresponding palette is shown, replacing the standard code completion menu. Figure 1a gives an example of a simple palette that may be associated with the \verb|Color| class.

The developer can interact with such palettes to provide parameters and other information related to her intent, and receive immediate feedback about the effect these choices will have on the behavior of the object being constructed. When this interaction is complete, the palette is responsible for generating appropriate source code for insertion at the cursor. Figure 1b shows the code that is inserted after the user presses the \textsc{enter} key.

In accordance with rational design practices, we sought to address the following questions before designing and implementing our active code completion system:

\begin{itemize}
\item What {\it specific} use cases exist for this form of active code completion in a professional development setting? 
\item What {\it general} criteria are common to types that would and would not benefit from an associated palette?
\item What are some relevant usability and design criteria for palettes designed to address such use cases?
\item What capabilities must the underlying active code completion system provide to enable these use cases and user interface designs?
\end{itemize}

To help us answer these questions, we conducted a large online survey of professional developers (Section 2). Their responses, along with information gathered from informal interviews and code corpus analyses, revealed a number of interesting use cases (Section 3) and non-trivial functional requirements for palette interfaces as well as the underlying active code completion architecture (Section 4). Notably, many of these findings can be applied to development tools for professional developers more generally, not simply those integrated into the editor in the way we have shown.

Next, we describe Graphite, an Eclipse plug-in that provides active code completion for users of the Java programming language (Section 5). We describe several interesting design choices that were made to satisfy many of the requirements discovered in our preliminary investigations and briefly examine the trade-offs that were necessary due to Java's purely textual program representations and Eclipse's limited extension APIs.

Finally, we conduct a controlled lab study with a more complex palette, implemented using Graphite, that assists  developers as they write regular expressions (Section 6). The data and observations gathered from this study provide  specific evidence in support of the broader claim that highly-specialized tools that are integrated directly with the editing environment are particularly useful, and that an active code completion system like Graphite makes developing, deploying and discovering such tools fundamentally simpler.

\section{Survey Results}
To validate our general conceptualization of active code completion, develop concrete design criteria and create a concrete list of use cases, we began by conducting a large survey of professional software developers.

% In order to convey the active code completion concept clearly, we came up with three different use cases where we belive the active code completion technique will benefit developers frequently. For each use case, we designed three mockup screenshots and showed them in sequence: before invoking the palette, during the interaction with the palette, and after the palette is closed and code is generated. By doing this, the respondents could understand the concept precisely and give useful feedback relevant to the above questions.
\subsection{Survey Methods}
We recruit participants for this survey\footnote{https://www.surveymonkey.com/s/2GLZP8V} primarily from a popular programming-related discussion forum hosted on the popular website reddit.com\footnote{http://www.reddit.com/r/programming}. An additional 22 participants were recruited from the local computer science graduate student body.

Recruitment materials in both cases stated that we were seeking developers ``familiar with an object-oriented programming language like Java, C\# or Visual Basic and an integrated development environment like Eclipse or Visual Studio''.
Participants were told that the survey would take approximately 20 minutes to complete, and no reward was offered. Of the 696 people who started the survey, 473 participants (68\%) completed it. We examine the responses from completed surveys only in the analyses below.

\subsection{Familiarity with Programming Languages and Editors}

We first asked participants about their level of familiarity with several programming languages, on a five-point Likert scale ranging from ``None'' to ``Expert''. 61.1\% of the participants indicated that they were an expert in at least one language, and an additional 35.7\% were ``very familiar'' with at least one language. On average, participants rated themselves as very familiar with Java, C, C++ and JavaScript, familiar with C\#, Python and PHP and somewhat familiar with Visual Basic and Perl.

We also asked participants to select which integrated development environments (IDEs) and code editors that they were familiar with. The Eclipse IDE was familiar to 87.1\% of participants. This was followed by Visual Studio at 66.0\%, Vi/Vim at 53.7\%, Netbeans at 37.7\%, Emacs at 24.8\% and IntelliJ IDEA at 16.4\%. Participants could also enter ``other'' choices and a number of editors and IDEs were entered, including Xcode, Textmate and Notepad++.

\subsection{Palette Mockups}

%\begin{figure}
%  \includegraphics[width=\linewidth]{mockup-palette-small.png}
%  \caption{An example mockup palette for regular expressions shown to the survey participants. In the survey, they were shown the surrounding Eclipse IDE together.}
%  \label{regex}
%\end{figure}
%
Next, we presented participants with a series of mockup palettes for a Color class (more complex than the one we ultimately implemented in Figure 1a), a regular expression class, and a SQL query class. Participants were also shown mockup screenshots demonstrating how a user could invoke the palette, and a mockup showing the code that would be inserted once a selection had been made. Before presenting each mockup, we gathered information about  the strategies that they would normally use to instantiate the class.

For the Color class, the majority of participants indicated that they would look in the code completion menu (58.4\%) or in the class documentation (19.0\%) for a predefined constant if asked to instantiate an object corresponding to the color ``navy'' (which is not, in fact, a standard color in Java.) Another 14.0\% indicated that they would use an external tool (such as an image editor) to determine the RGB values corresponding to the color.

Before asking participants about regular expressions and SQL queries, we asked participants to rate their familiarity with these concepts. Few participants (4\%) indicated that they were unfamiliar with regular expressions and no participants were unfamiliar with creating SQL queries, providing further evidence that our participants were not novice developers. Fig. \ref{strategies} summarizes the strategies that participants generally preferred for instantiating regular expressions and SQL queries. Using an external tool was a common strategy in both cases, particularly for SQL queries, but several other strategies were also represented.

\begin{figure}
\begin{tabular}{lccc}
 & Regular Expressions &  & SQL\\
 \hline
Separate test script & 29.6\% &  & 15.4\%\\
Guess and check & 14.0\% &  & 16.1\%\\
External tool & \textbf{37.9\%} &  & \textbf{58.6\%}\\
Search for examples & 12.3\% & & 5.1\% \\
Other & 6.2\% & & 4.9\% \\
\hline
\end{tabular}
\caption{Distribution of responses to survey questions asking about typical strategies for writing regular expressions and SQL queries.}
\label{strategies}
\end{figure}

\begin{figure}
\begin{tabular}{crccccc}\\\\
\textsc{class}
& 
& \begin{rotate}{20}Nearly every time\end{rotate}
& \begin{rotate}{20}Most of the time\end{rotate}
& \begin{rotate}{20}Some of the time\end{rotate}
& \begin{rotate}{20}Rarely\end{rotate}
& \begin{turn}{20}Never\end{turn}\\
\hline
Color &\vline& 9.6\% & 22.1\% & \textbf{32.4\%} & 28.2\% & 7.7\%\\
RegExp &\vline& \textbf{36.6\%} & 29.5\% & 21.8\% & 7.3\% & 4.8\%\\
SQL & \vline &18.2\% & 19.3\% & \textbf{30.9\%} & 20.4\% & 11.4\%\\
\hline
\end{tabular}
\caption{The distribution of responses to the question: ``Consider situations where you need to instantiate the [specified] class. What portion of the time, in these situations, do you think you would use this feature?''}
\label{frequencies}
\end{figure}

Finally, after showing the series of mockup screenshots, we asked participants to rate how useful the integrated palette would be to them if they needed to instantiate the corresponding class. The responses to this question for each palette are summarized in Fig. \ref{frequencies}. In each case, more than half of the participants indicated that they would use active code completion at least some of the time. The regular expression palette was considered particularly useful while the color and SQL palettes showed a more reserved pattern of responses.

In addition to asking for a simple rating for each palette, we also solicited open-ended comments. A large number of participants volunteered comments: 193 for the color palette, 129 for the regular expression palette and 142 for the SQL palette. These responses were highly valuable when developing the design criteria below and helped to explain the patterns observed in Fig. \ref{frequencies}.

\section{Design Criteria}
Using the information gathered from the survey as well as informal discussions with developers and researchers over the course of this study, we sought to develop design criteria constraining both the overall system design as well as the design of individual palettes. In the section headings below, the number of survey responses, summed over the three palette mockups, that contained the listed concern, as judged by the authors of this paper, are listed in parenthesis. Many of these concerns apply generally to various kinds of editor-integrated tools, not just those implemented using the active code completion mechanism that we focus on, so we hope these design criteria will be valuable to the developer tool design community more broadly.

\subsection{Maintaining Separation of Concerns (183)}
The most common issue developers had, particularly with the Color palette, was that it seemed to ignore the issue of separation of concerns by allowing developers to insert color constants directly into the program logic. Many developers noted that this is considered bad practice, or should be limited to the prototyping phase of a project. This concern was also expressed in responses to the SQL palette, which required inserting parameters to connect to a database so that the query could be tested. The code that was inserted included initialization steps needed to connect to the specific database that was entered, and several participants noted that much of this information should appear in an external resource file rather than inline with the program logic itself. Few participants made similar comments about the regular expression palette, however, indicating that regular expressions are considered a part of the program logic rather than data by most developers.

This feedback suggests that tool and palette designers may wish to explicitly aprise users of relevant best practices and acknowledge that palettes that generate constant data may be most useful in the prototyping phase. It can be noted that when transitioning from a prototype to production-quality software the code generated by a palette or tool may be used as template that can be refactored to follow stricter policies as needed. We showed users the Color and SQL palettes in the context of a standard programming editor, but these comments indicate that resource file and  stylesheet editors may benefit from active code completion support as well.

\subsection{Integration with Testing Frameworks (35)}
The regular expression palette shown to the participants allowed users to immediately test a pattern against provided strings. These test strings and the results of performing the match were inserted as comments below the generated source code. A number of participants requested that explicit unit tests be generated instead. Although the palette interface conveyed the results of these tests immediately, participants remained adamant about explicit unit tests, likely due to concerns that future modifications to the generated code without reinvoking the palette might introduce bugs, or due to the desire to conform to standard testing practices. To support the generation of unit tests, however, the active code completion architecture must explicitly allow for code generation at locations other than at the cursor and handle the fact that statements generated after a code completion command are not necessarily encapsulated well enough to be immediately testable in isolation.

%Examples of feedbacks in this category include:
%
%\begin{itemize}
%	\item The use of magic numbers in the source code is a bad design (54)
%	\item Color picking is the `designer's job' (33)
%	\item Would only use to define named constants (17)
%	\item Would rather use color scheme / theme (3)
%	\item Would use Object-Relational Mapping (ORM) / Language INtegrated Query (LINQ) instead of raw SQL query string (27 of 142)
%	\item These palettes might be useful for prototyping, but not for real production (14)
%	\item Instead of including the test cases into comments, they should be stored as unit tests (35 of 143)
%\end{itemize}

\subsection{Support for Reinvocation (19)}
Several participants asked for the ability to reinvoke a palette from previously generated source code. In order to support this feature, the architecture must provide the palette with enough information to reconstruct its state. To complicate matters, however, users may wish to modify the generated code between invocations of the palette and have these modifications reflected in the palette's state upon reinvocation (e.g. modify the RGB values in the case of the \verb|Color| class). Moreover, there may be important aspects of the palette's state that are not directly available in the generated code, such as parameters controlling the palette's user interface. Indeed, associating tool-related metadata with code is known to be cumbersome in purely textual languages, since all metadata must be directly visible within comments or annotations.

\subsection{Support for Palette Settings and History (41)}

A related feature important to many participants is support for maintaining settings and usage history across invocations of the same palette at different code locations. For example, 20 participants requested that the Color palette include a list of recent or favorite colors, and 12 participants specifically inquired about whether the database connection information was maintained between invocations of the SQL palette.

% The followings are the comments that are related to this category.

%\begin{itemize}
%	\item Want recently used regular expressions, and control over the comments (12)
%	\item Persistent database connection information is needed / want to manage connection pools (9)
%	\item Want recent colors, capability of setting favorite colors (20)
%\end{itemize}

\subsection{Support for Nested Expressions (13)}

In all of the examples that we gave, the parameters entered into the palette interface corresponded to simple, constant expressions, rather than complex expressions referring to variables from the surrounding context. A number of participants noticed this limitation. For example, several participants asked SQL query strings, as these are typically constructed using user-generated data in practice. Although a simple expression entry box may suffice in simple scenarios, architectural support is needed for palettes that need to inspect the code context (e.g. to verify well-formedness) or if code highlighting, code completion and other advanced editing features are needed within the palette itself.

\subsection{Keyboard Navigability (12)}

Although our presentation of active code completion did not include any completely mouse-driven interfaces, several participants commented that the Color palette included interface elements taken from standard color dialog boxes that could only be manipulated using the mouse. These comments were generally severe in their condemnation of mouse-based interfaces in developer tools, consistent with our finding that a significant portion of our participants were familiar with editors like Vim that place a strong emphasis on keyboard shortcuts.

\subsection{Minimal Impact on Editor Responsiveness}

A common theme in our discussions with developers (as well as in comments left on our recruitment thread) was that integrated development environments like Eclipse were already too slow, and that an extension such as the one we were proposing would only be acceptable if it did not affect performance and responsiveness any further.

\subsection{IDE and Language Portability}

The mockups we showed users were based on the Java language and the Eclipse development environment. As we showed, a number of participants preferred other languages or editors. Many of these participants made comments asking that the features we describe not be restricted to one specific editor (e.g. NetBeans users) and some also called for support for other programming languages (e.g. Ruby and C\#). Indeed, the palettes we demonstrated could be used with only slight modifications in a variety of programming languages, given suitable architectural support for porting palettes between editing environments.

\subsection{Varying User Needs}

Our initial color palette was deemed overly complex by many participants (26), with many others (26) specifically saying that they would be satisfied with a simpler interface with colors listed next to their names. Other users wanted additional features, such as an ``eyedropper'' tool for selecting a color directly from an image on the screen. Similarly, the regular expression and SQL palettes were considered too simple by 12, and 15 participants respectively. User requests included syntax highlighting, a range of additional helpful tools for generating,  testing and sharing regular expressions, and the ability to browse databases more directly. Due to the wide range of user needs even within a single class, it may be that support for multiple palettes or a tabbed palette interface will be necessary in practice. To support incremental improvements to palettes based on user feedback, an architecture that makes deploying new and updated palettes relatively painless would also be valuable.

\section{Use Cases}
At the end of the survey, we solicited suggestions from the participants of other classes that could benefit from an associated palette. 119 participants made one or more suggestion and we classified the suggestions into several categories. We also mention other suggestions made by researchers and developers via other channels in some cases, without including them in the provided counts.

\subsection{Graphical Elements (27)}
The most popular suggestions were graphical elements, influenced perhaps by our demonstration of the Color palette. Some participants suggested palettes for classes representing primitive graphical objects, such as brush and font selectors or polygon editors, while other participants were focused on user interface elements, such as buttons, check boxes and frame layouts. A few also suggested palettes for manipulating 3D primitives, such as transformation matrices, in a more direct and intuitive manner. A practitioner also suggested that because setting up a plot or graph is often significantly simpler using a direct manipulation interface, it would be a natural candidate for a palette as well.

\subsection{Query Languages (17)}
The second most popular category of suggestions consisted of various interfaces for query languages, also likely due to the examples we provided to participants. In addition to variants of the SQL and regular expression palettes, developers also wanted to work with other types of queries such as XPath or XQuery for XML.

\subsection{Simplified or Domain-Specific Syntax (16)}
Another interesting class of suggestions were cases where a more natural syntax than the syntax provided by Java is desirable. One suggestion was a palette that automatically escaped strings containing quotation marks or escape sequences. A related category of suggestions consisted of palettes that offered a more natural interface for generating strings containing code in other languages such as HTML (e.g. offering syntax highlighting, escaping,  tag matching and other features.) Domain-specific syntax for complex mathematical expressions and chemical formula were also mentioned in discussions with practitioners.

An interesting suggestion that we investigated further involved Java's collection classes, such as \verb|ArrayList| and \verb|HashMap|. A participant suggested that these classes could be associated with a palette that offered a simplified literal syntax for initialization, pointing toward other languages that do offer such a literal syntax (e.g. JavaScript.) Without such syntax, these classes must be tediously initialized using a separate method call for each element. To determine whether this usage pattern is common, we conducted a corpus analysis using 10 randomly selected projects from the Qualitas Corpus \cite{QualitasCorpus:APSEC:2010} containing over 1M lines of code. We began by searching for places in these projects where Java collection classes were being instantiated, then looked to see whether this instantiation code was immediately followed by method calls that inserted items into the collection, indicating a case where a literal may have been used if available. Figure 4 summarizes the results of this analysis, providing evidence in support of the claim that a palette that simplifies this process could be useful for general-purpose programming.

\subsection{Unclear Parameter Implications (11)}
Another category of use cases contains classes where it can be difficult to predict what the run-time behavior of a particular parameter choice may be. Examples given included audio filters (e.g. pitch manipulation) and animation descriptors (e.g. speed or shape parameters). By giving immediate visual or auditory feedback using a preview panel, these parameters can be tweaked without requiring the execution of the full application.

%
%\begin{itemize}
%	\item Example classes
%	
%	\begin{itemize}
%		\item String
%		\item Collections (e.g., dictionary)
%		\item Vector, Matrix
%		\item Embedded languages
%		\item URLs, Paths
%	\end{itemize}
%	
%	\item Expected features
%	
%	\begin{itemize}
%		\item Proper escaping
%		\item Syntax highlighting
%		\item Auto-completion
%	\end{itemize}
%\end{itemize}

\begin{figure}\label{collections}
\begin{center}
\begin{tabular}{l|c|c|r}
Collection Class& Total & Literal & Percentage \\
\hline
\texttt{ArrayList} & 464	&	44	&	9.5\% \\
\texttt{HashMap}	& 56 & 19 & 33.9\% \\
\texttt{HashSet} & 122 & 62 & 50.8\% \\
\texttt{Hashtable} & 86 & 10 & 11.6\% \\
\texttt{Vector} & 729 & 31 & 4.2\% \\
\hline
\textbf{Total} & 1457 & 166 & 11.4\% \\
\hline
\end{tabular}
\caption{Usage patterns for common Java collection classes in the \texttt{java.util} package in our code corpus. Uses that fit a pattern that can be captured by a literal make up a significant portion of all uses. Not all possible usage scenarios of this type were captured by our analysis, so these numbers are lower bounds.}
\end{center}
\end{figure}

%\begin{itemize}
%	\item Example classes
%	
%	\begin{itemize}
%		\item Audio tweaks (modifying pitch, volume, etc.)
%		\item 3D transformation matrices
%		\item number / string / date formatting
%		\item message box / input box (e.g., \texttt{JOptionPane})
%	\end{itemize}
%\end{itemize}

%\begin{itemize}
%	\item Example classes
%	
%	\begin{itemize}
%		\item Regular expression
%		\item SQL query
%		\item XPath / XQuery
%	\end{itemize}
%\end{itemize}

%\begin{itemize}
%	\item Example classes
%	
%	\begin{itemize}
%		\item Color, Shape, Brush, Font, etc.
%		\item JFrame, Layouts, any swing components
%	\end{itemize}
%\end{itemize}

\subsection{Integrating with Documentation and Examples (7)}
Some participants suggested integrating tutorials or lists of relevant examples directly using a palette, so that these can be discovered more easily by new users and inserted directly into code, without requiring switching to a web browser and executing a search.

\subsection{Complex Instantiation and Cleanup Procedures (5)}
A related category contains classes that require complex instantiation and cleanup procedures. For example, in order to read a text file in Java, the developer might want to use \texttt{BufferedReader} class. This class can be difficult to use because it requires try/catch block, and one must remember to close the file after reading it. By using a palette to choose a file or choose a variable which contains the file path, the developers could easily instantiate these objects and get an outline containing the full life-cycle of the file. Similarly, palettes may help to alleviate the factory pattern usability problem \cite{ellis_factory_2007}. As long as the developers remember which class to use, they will not need to remember how to instantiate that class. We explore this further in our user study in Section VI.

%\begin{verbatim}
%BufferedReader reader = null;
%try {
%    reader = new BufferedReader(
%        new FileReader(filePath));
%} catch (IOException e) {
%    e.printStackTrace();
%}
%// use reader here
%reader.close();
%\end{verbatim}

\subsection{Instantiation by Example (2)}
In some cases, it is possible to describe an object by example. For instance, a class that represents a shortcut key combination may be most easily instantiated using a palette that simply reads a shortcut key from the developer.

\subsection{Proof Assistants (1)}
A proof assistant is a tool for constructing proof terms. According to the well-known Curry-Howard isomorphism between programming languages and formal logics, proof terms correspond to expressions and propositions correspond to types \cite{tapl}. Active code completion works directly with types to help developers construct expressions, so if applied to a language with a cleanly-developed connection to formal logic (e.g. Coq), palettes would be useful for constructing highly interactive proof assistant interfaces.

%\begin{itemize}
%	\item Example classes
%	
%	\begin{itemize}
%		\item Keyboard Keys
%		\item Regular expression
%	\end{itemize}
%\end{itemize}

%\section{Tool}
%Described earlier.
%\section{Developer Survey}
%
%\subsection{Methods}
%Our subject pool consisted primarily of visitors to a large collaborative filtering website dedicated to programming called {\tt /r/programming} on reddit.com\footnote{http://www.reddit.com/r/programming}. At the time of our study, there were approximately 340,000 registered members. A small number of participants were also recruited using a mass email to local computer science graduate students. The recruitment materials in both cases stated that we were seeking developers ''familiar with an object-oriented programming language like Java, C\# or Visual Basic and an integrated development environment like Eclipse or Visual Studio''.
%
%Participants were asked to take an online survey that would take approximately 20 minutes to complete. Of the 696 participants who started the survey, 475 participants finished it. Of these, 457 participants came from /r/programming and 18 from the graduate student mailing list. All quantitative metrics used in this study are based only on these participants.
%
%\subsection{Familiarity with Languages and Editors}
%
%We began by asking participants to rate their level of familiarity with several programming languages. Figure 1 summarizes these ratings. On average, participants rated themselves as very familiar with Java, C, C++ and JavaScript, familiar with C\#, Python and PHP and somewhat familiar with Visual Basic and Perl.
%
%We also asked participants to select the integrated development environments and code editors that they were familiar with. The Eclipse IDE was familiar to 87.1\% of participants. This was followed by Visual Studio with 66.0\% of participants, Vi/Vim with 53.7\%, Netbeans with 37.7\%, Emacs with 24.8\% and IntelliJ IDEA with 16.4\%. Participants could also enter ``other'' choices. A number of editors and IDEs were entered, including Xcode, Textmate, Notepad++ and others.
%%\begin{figure}
%%  \includegraphics[width=3.5in,height=2in]{experience.png}
%%  \caption{Participant familiarity with various programming languages. Starting at bottom: None, Somewhat familiar, Familiar, Very Familiar, Expert.}
%%\end{figure}
%
%\subsection{Mockups}
%%\begin{figure*}
%%  \includegraphics[scale=.6]{mockup-palette.png}
%%  \caption{The regular expression palette, shown in Eclipse.}
%%\end{figure*}
%Next, we presented participants with a series of mockup palettes corresponding to a Color class, a regular expression Pattern class (Figure 2) and a SQL query ResultSet class. In addition to a screenshot of the palette itself, participants were shown mockups demonstrating how a user would invoke the palette, and a mockup showing the code that would be inserted once a selection had been made. 
%
%Before presenting each palette mockup we gathered information about the user's familiarity with that topic and the strategies that they would likely use to instantiate that class. The Color class was presumed familiar to all participants. Figure 3 summarizes responses for the regular expression and SQL classes. The low number of participants unfamiliar with these topics again provides evidence that our participants were not generally novice developers.
%
%To determine the strategy typically used for instantiating the Color class, we asked the following question. The values in bold show the frequency of each response.
%
%\begin{quote}
%A user interface designer asks you to change the background color of a window to ``navy blue'' when the user presses a button. You have determined that you need to instantiate the 'Color' class to do so.
%\\\\
%Which of the following strategies would you use first to create an instance of the 'Color' class representing navy blue?\\
%
%\begin{enumerate}[(a)]
%\item I would use the IDE's content assist menu (see note below) to look for a predefined constant representing navy blue. (\textbf{58.4\%})
%\item I would look up the documentation of the Color class to see if there is a predefined constant representing navy blue. (\textbf{19.0\%})
%\item I would use an external tool (e.g. Photoshop or some other image editor) to determine the RGB values for navy blue. (\textbf{14.0\%})
%\item I would guess at the RGB values for navy blue, then check whether my guess was correct by executing the application. (\textbf{5.3\%})
%\item Other (please specify) (\textbf{3.3\%})\\
%\end{enumerate}
%\end{quote}
%
%Figure 4 summarizes the responses to a more general question asked about regular expressions and SQL queries.
%
%\begin{figure}
%\begin{tabular}{lccc}
% & \textbf{Regular Expressions} &  & \textbf{SQL}\\
% \hline
%Never used & 4.8\% &  & 0.0\%\\
%Use infrequently & 46.7\% &  & 37.4\%\\
%Use frequently & \textbf{48.4\%} &  & \textbf{62.6\%}\\
%\hline
%\end{tabular}
%\caption{Participant's experience with regular expressions and SQL.}
%\end{figure}
%
%\begin{figure}
%\begin{tabular}{lccc}
% & \textbf{Regular Expressions} &  & \textbf{SQL}\\
% \hline
%Separate test script & 29.7\% &  & 15.5\%\\
%Guess and check & 13.4\% &  & 16.0\%\\
%External tool & \textbf{38.5\%} &  & \textbf{58.9\%}\\
%Search for examples & 12.1\% & & 4.8\% \\
%Other & 6.2\% & & 4.8\% \\
%\hline
%\end{tabular}
%\caption{Typical strategies for regular expressions and SQL queries.}
%\end{figure}
%
%
%%For each of these classes,
%
%
%%\subsubsection{Color}
%%We began with a class representing colors. We asked participants to consider the following scenario:
%
%\begin{figure}
%\begin{tabular}{crccccc}\\\\
%\textsc{class}
%& 
%& \begin{rotate}{20}Nearly every time\end{rotate}
%& \begin{rotate}{20}Most of the time\end{rotate}
%& \begin{rotate}{20}Some of the time\end{rotate}
%& \begin{rotate}{20}Rarely\end{rotate}
%& \begin{turn}{20}Never\end{turn}\\
%\hline
%Color &\vline& 9.6\% & 22.1\% & \textbf{32.4\%} & 28.2\% & 7.7\%\\
%RegExp &\vline& \textbf{36.6\%} & 29.5\% & 21.8\% & 7.3\% & 4.8\%\\
%SQL & \vline &18.2\% & 19.3\% & \textbf{30.9\%} & 20.4\% & 11.4\%\\
%\hline
%\end{tabular}
%\caption{For each of the classes considered, participants were shown mockups of a code completion palette designed for instantiating that class, as well as the source code generated after parameters were entered. This table shows the distribution of responses to the question: ``Consider situations where you need to instantiate the [specified] class. What portion of the time, in these situations, do you think you would use this feature?''}
%\end{figure}
%%
%%\begin{figure}
%%  \includegraphics[scale=.6]{regex-insert.png}
%%  \caption{Test}
%%\end{figure}
%%
%%
%
%%Then we asked:
%%\begin{quote}
%%\textbf{Q2} Consider situations where you need to instantiate the 'Color' class. What portion of the time, in these situations, do you think you would use this feature?
%%\begin{enumerate}[(a)]
%%\item Nearly every time (\textbf{9.6\%})
%%\item Most of the time (\textbf{22.1\%})
%%\item Some of the time (\textbf{32.4\%})
%%\item Rarely (\textbf{28.2\%})
%%\item Never (\textbf{7.7\%})\\
%%\end{enumerate}
%%\end{quote}
%%
%
%After presenting each palette, we asked users to rate the usefulness of each palette. Responses are summarized in Figure 5. We then allowed participants to provide qualifications and open-ended feedback by asking the following question: ``If you would like to qualify or elaborate on your answer to the previous question you may do so below. You may also suggest improvements for this palette.'' This information is analyzed in the remainder of the paper.
%
%%\begin{quote}
%%\textbf{Q4} How would you describe your level of familiarity with (regular expressions, SQL)?\\
%%\\
%%\begin{enumerate}[(a)]
%%\item I do not know what [it is].	 (\textbf{0.7\%}, \textbf{0.0\%})
%%\item I know what regular expressions are but I have not used them in practice. (\textbf{4.4\%, 0.0\%})
%%\item I use regular expressions infrequently.	 (\textbf{46.0\%, 35.9\%})
%%\item I use regular expressions frequently. (\textbf{49.0\%, 64.1\%})
%%\end{enumerate}
%%\end{quote}
%
%
%
%%\begin{quote}
%%\textbf{Q5} Consider situations where you need to write a regular expression to extract information from a string or collection of strings. What strategy would you most likely use first?
%
%%\begin{enumerate}[(a)]
%%\item I would write a separate test script to help come up with an appropriate regular expression. (\textbf{29.7\%})
%%\item I would guess and check regular expressions by running my application. (\textbf{13.4\%})
%%\item I would launch an external application or web tool to help me test regular expressions interactively. (\textbf{38.5\%})
%%\item I would search the web for a sample regular expression suitable for my task. (\textbf{12.1\%})
%%\item Other (please specify) (\textbf{6.2\%})
%%\end{enumerate}
%%\end{quote}
%
%
%%\begin{quote}
%%\textbf{Q6} Consider situations where you need to create a regular expression. What portion of the time, in these situations, do you think you would use this feature?
%%\begin{enumerate}[(a)]
%%\item Nearly every time (\textbf{36.6\%})
%%\item Most of the time (\textbf{29.5\%})
%%\item Some of the time (\textbf{21.8\%})
%%\item Rarely (\textbf{7.3\%})
%%\item Never (\textbf{4.8\%})\\
%%\end{enumerate}
%%\end{quote}
%
%%\begin{quote}
%%\textbf{Q7} How would you describe your level of familiarity with SQL?
%
%%\begin{enumerate}[(a)]
%%\item I do not know what regular expressions are.	 (\textbf{0.0\%})
%%\item I know what SQL is I have not used it in practice. (\textbf{0.0\%})
%%\item I use SQL infrequently.	 (\textbf{35.9\%})
%%\item I use SQL frequently. (\textbf{64.1\%})\\
%%\end{enumerate}
%%\end{quote}
%
%%\begin{quote}
%%\textbf{Q8} Consider situations where you need to write a regular expression to extract information from a string or collection of strings. What strategy would you most likely use first?
%
%%\begin{enumerate}[(a)]
%%\item I would write a separate test script to help come up with an appropriate SQL query.	
% %(\textbf{15.5\%})
%%\item I would guess and check by running my application. (\textbf{16.0\%})
%%\item I would launch an external application or web tool to test my query against my database interactively.	(\textbf{58.9\%})
%%\item I would search the web for an example SQL query suitable for my task.	
% %(\textbf{4.8\%})
%%\item Other (please specify) (\textbf{4.8\%})
%%\end{enumerate}
%%\end{quote}
%
%%Showed the SQL palette and asked:
%
%%\begin{quote}
%%\textbf{Q9} Consider situations where you need to create a SQL query. How often, in these situations, do you think you would use this feature?
%%\end{quote}
%
%\section{Design Implications}
%%\begin{figure}[b]
%%  \includegraphics[scale=.5]{color-palette.png}
%%  \caption{test}
%%\end{figure}
%
%We next analyzed the open-ended responses to extract some design principles and constraints. We got 193 responses for the color palette, 129 for the regular expression palette, and 142 for the SQL palette. Also, we got total 119 suggestions about what other types of classes could benefit from similar types of palettes. The design can be divided into two main categories: system design constraints, and general palette design considerations. System design constraints are the criteria that should be taken into account in the design of the overall system architecture, and the general palette design considerations are about how to make more usable palettes.
%
%\subsection{System Design Constraints}
%
%\subsubsection{Handling Separation of Concerns}
%
%Professional developers were very wary of including constant data into the source code directly. Many responses expressed general sentiment or specific preference for project-wide color theme or external resource file. Also, many people stated that these tools might not be useful since they usually read an external resource file, or sometimes they explicitly suggested to add capability to handle the external resource files directly from the palette.
%
%Examples of feedbacks in this category include:
%
%\begin{itemize}
%	\item The use of magic numbers in the source code is a bad design (54)
%	\item Color picking is the `designer's job' (33)
%	\item Would only use to define named constants (17)
%	\item Would rather use color scheme / theme (3)
%	\item Would use Object-Relational Mapping (ORM) / Language INtegrated Query (LINQ) instead of raw SQL query string (27 of 142)
%	\item These palettes might be useful for prototyping, but not for real production (14)
%	\item Instead of including the test cases into comments, they should be stored as unit tests (35 of 143)
%\end{itemize}
%
%Interestingly, most people seemed to consider regular expressions as part of the program logic. There were few people who complained about including the regular expression pattern string in the source code.
% 
%\subsubsection{Reversibility}
%
%19 participants expressed that it should be able to bring the palettes back when they want to modify the parameters after they used the palettes.
% 
%\subsubsection{Palette Settings and State}
%
%Many participants wanted to be able to configure the palettes or have the palettes maintain state even after the palettes are hidden. The followings are the comments that are related to this category.
%
%\begin{itemize}
%	\item Want recently used regular expressions, and control over the comments (12)
%	\item Persistent database connection information is needed / want to manage connection pools (9)
%	\item Want recent colors, capability of setting favorite colors (20)
%\end{itemize}
%
%\subsubsection{Interaction with Code Context}
%
%Some people wanted to interact with the code context. For example, 13 people expressed that it would be great if the SQL query palette was capable of dynamically creating SQL query string by combining multiple string variables, or by assigning variables for the parameters.
%
%\subsubsection{Performance}
%
%Developers were concerned about the performance of these palettes. Some of the participants said that they would use these palettes only if it does not slow down the IDE. Also, this was the most popular comment on the reddit thread.
%
%\subsubsection{IDE Independence}
%
%Several participants expressed a desire for IDE or even language independence for these palettes.
%
%%\subsubsection{Composing palette logic}
%
%\subsection{General Palette Design Considerations}
%\subsubsection{Simplicity vs. Capability}
%
%Even though the participants were shown the same mockup screenshots, their reactions were split. Our color palette was considered too complex by many participants (26), and many others (26) would be satisfied with seeing colors next to a list of color names. The regular expressoin palette and SQL palette were considered too simple (12, and 15 participants respectively). They wanted syntax highlighting, match highlighting, and browsing capability of SQL databases.
%
%\subsubsection{Keyboard Navigability}
%
%12 of the participants expressed strong antipathy toward the mouse. This was mostly directed at the color palette.
%
%\subsubsection{Complexity}
%
%%\subsubsection{Side Effects}
%
%\section{Suggestions from the Participants}
%We classified the suggestions into several categories. The number in the parentheses indicate the number of participants who suggested the classes which fall into the category.
%
%\subsection{Alternative/Tricky/Literal Syntax (16)}
%These are the classes of which syntax is tricky, or complex. Even for a basic string class, people wanted to input multiline string or unicode string in more convenient way with automatic escaping.
%
%\begin{itemize}
%	\item Example classes
%	
%	\begin{itemize}
%		\item String
%		\item Collections (e.g., dictionary)
%		\item Vector, Matrix
%		\item Embedded languages
%		\item URLs, Paths
%	\end{itemize}
%	
%	\item Expected features
%	
%	\begin{itemize}
%		\item Proper escaping
%		\item Syntax highlighting
%		\item Auto-completion
%	\end{itemize}
%\end{itemize}
%
%\subsection{Unclear Parameter Implications (11)}
%The classes in this category has some parameters in it, and it is not easy to predict the results of the parameter modifications. Instead of running the whole application to see the result, the palette can serve as a preview panel so that the developers may check the results directly. Also, the palette can be a control panel which facilitates modifying the parameters.
%\begin{itemize}
%	\item Example classes
%	
%	\begin{itemize}
%		\item Audio tweaks (modifying pitch, volume, etc.)
%		\item 3D transformation matrices
%		\item number / string / date formatting
%		\item message box / input box (e.g., \texttt{JOptionPane})
%	\end{itemize}
%\end{itemize}
%
%\subsection{Query Languages (17)}
%Similar to the unclear parameter implications category, the developers wanted to check if the query string is correct or not by checking the results directly on the palette.
%
%\begin{itemize}
%	\item Example classes
%	
%	\begin{itemize}
%		\item Regular expression
%		\item SQL query
%		\item XPath / XQuery
%	\end{itemize}
%\end{itemize}
%
%\subsection{Graphical Elements (27)}
%The other popular category was graphical elements. We've already shown the color example, and the participants suggested many other graphical properties such as font, brush, shape and others. Also, people wanted to check or directly manipulate the GUI frames and layouts.
%
%\begin{itemize}
%	\item Example classes
%	
%	\begin{itemize}
%		\item Color, Shape, Brush, Font, etc.
%		\item JFrame, Layouts, any swing components
%	\end{itemize}
%\end{itemize}
%
%\subsection{Complex Instantiation Procedures (5)}
%Some of the participants pointed out that these palettes might be useful to create an object which requires complex instantiation code. For example, in order to read a text file in Java, the developer might want to use \texttt{BufferedReader} class. The instantiation code might look as following:
%\begin{verbatim}
%BufferedReader reader = null;
%try {
%    reader = new BufferedReader(
%        new FileReader(filePath));
%} catch (IOException e) {
%    e.printStackTrace();
%}
%// use reader here
%reader.close();
%\end{verbatim}
%By using a palette to choose a file or choose a variable which contains the file path, the developers could easily instantiate these objects. In the same way, it can alleviate the factory pattern usability problem. As long as the developers remember which class to use, they will not need to remember how to instantiate that class.
%
%\begin{itemize}
%	\item Example classes
%	
%	\begin{itemize}
%		\item BufferedReader
%		\item Classes using factory patterns
%		\item Database connection string builder
%	\end{itemize}
%\end{itemize}
%
%\subsection{Describe by Example (2)}
%Sometimes, it is possible to describe an object by examples. For instance, if there is a class which represents a shortcut key combination, we can easily instantiate an object of that class by pressing the actual shortcut key on the interactive palette.
%
%\begin{itemize}
%	\item Example classes
%	
%	\begin{itemize}
%		\item Keyboard Keys
%		\item Regular expression
%	\end{itemize}
%\end{itemize}
%
%\subsection{Integrating with Documentation, Tutorial (7)}
%Some participants suggested integrating the documentation or the tutorials into the interactive palettes. If the developers are not familiar of the specific API class, the palette could tell them how to use it.
%
%%\begin{itemize}
%%\item 54 expressed a concern about encouraging inlining of constants, preferring external resource files. 17 stated that they would use this feature only to define constants. 33 stated that it should be the designer's job.
%%\item 11 expressed a concern about keyboard navigability
%%\item 27 expressed a concern about the complexity of the palette
%%\item 10 expressed an interest in a color picker. 7 wanted a multitabbed window with different options for picking colors.
%%\item 26 expressed a desire for color names, 10 of which wanted to ensure color constant names were used if available
%%\item 20 wanted persistent state (MRU, favorites)
%%\end{itemize}
%
%% An example of a floating figure using the graphicx package.
%% Note that \label must occur AFTER (or within) \caption.
%% For figures, \caption should occur after the \includegraphics.
%% Note that IEEEtran v1.7 and later has special internal code that
%% is designed to preserve the operation of \label within \caption
%% even when the captionsoff option is in effect. However, because
%% of issues like this, it may be the safest practice to put all your
%% \label just after \caption rather than within \caption{}.
%%
%% Reminder: the "draftcls" or "draftclsnofoot", not "draft", class
%% option should be used if it is desired that the figures are to be
%% displayed while in draft mode.
%%
%%\begin{figure}[!t]
%%\centering
%%\includegraphics[width=2.5in]{myfigure}
%% where an .eps filename suffix will be assumed under latex, 
%% and a .pdf suffix will be assumed for pdflatex; or what has been declared
%% via \DeclareGraphicsExtensions.
%%\caption{Simulation Results}
%%\label{fig_sim}
%%\end{figure}
%
%% Note that IEEE typically puts floats only at the top, even when this
%% results in a large percentage of a column being occupied by floats.
%
%
%% An example of a double column floating figure using two subfigures.
%% (The subfig.sty package must be loaded for this to work.)
%% The subfigure \label commands are set within each subfloat command, the
%% \label for the overall figure must come after \caption.
%% \hfil must be used as a separator to get equal spacing.
%% The subfigure.sty package works much the same way, except \subfigure is
%% used instead of \subfloat.
%%
%%\begin{figure*}[!t]
%%\centerline{\subfloat[Case I]\includegraphics[width=2.5in]{subfigcase1}%
%%\label{fig_first_case}}
%%\hfil
%%\subfloat[Case II]{\includegraphics[width=2.5in]{subfigcase2}%
%%\label{fig_second_case}}}
%%\caption{Simulation results}
%%\label{fig_sim}
%%\end{figure*}
%%
%% Note that often IEEE papers with subfigures do not employ subfigure
%% captions (using the optional argument to \subfloat), but instead will
%% reference/describe all of them (a), (b), etc., within the main caption.
%
%
%% An example of a floating table. Note that, for IEEE style tables, the 
%% \caption command should come BEFORE the table. Table text will default to
%% \footnotesize as IEEE normally uses this smaller font for tables.
%% The \label must come after \caption as always.
%%
%%\begin{table}[!t]
%%% increase table row spacing, adjust to taste
%%\renewcommand{\arraystretch}{1.3}
%% if using array.sty, it might be a good idea to tweak the value of
%% \extrarowheight as needed to properly center the text within the cells
%%\caption{An Example of a Table}
%%\label{table_example}
%%\centering
%%% Some packages, such as MDW tools, offer better commands for making tables
%%% than the plain LaTeX2e tabular which is used here.
%%\begin{tabular}{|c||c|}
%%\hline
%%One & Two\\
%%\hline
%%Three & Four\\
%%\hline
%%\end{tabular}
%%\end{table}
%
%
%% Note that IEEE does not put floats in the very first column - or typically
%% anywhere on the first page for that matter. Also, in-text middle ("here")
%% positioning is not used. Most IEEE journals/conferences use top floats
%% exclusively. Note that, LaTeX2e, unlike IEEE journals/conferences, places
%% footnotes above bottom floats. This can be corrected via the \fnbelowfloat
%% command of the stfloats package.
%
%
%
%\section{Conclusion}
%Overall, professional developers appear willing to use these for a number of use cases. However, the usability criteria described above impose stiff requirements on the architecture. Of particular concern, we found that the developers preferred help with program logic, but not the constant data. This indicates that another area of application may be in specialized editors for resource files.
%
%
\section{Tool Design}
Motivated by these use cases and with the design criteria developed as a result of our developer survey in mind, we sought to create an active code completion system. We named our project Graphite, an acronym for {\bf Gra}phical {\bf P}alettes to {\bf H}elp {\bf I}nstantiate {\bf T}ypes in the {\bf E}ditor. We describe the design decisions and trade-offs that went into this system in this section. A video demonstrating Graphite, including the process of writing a full palette, is available for further details \footnote{http://www.youtube.com/watch?v=FfCoXQEHGAI}. The full source code of Graphite is also available at the URL given on the first page.

We chose to use Eclipse for the Java programming language because this combination was the most widely-used  amongst participants in our survey. Eclipse plug-in development is well-supported and there are many examples available online that we can refer to, and the Eclipse project is available under an open source license.

\subsection{HTML+Javascript based palettes}
Eclipse is itself written in Java and uses the SWT and JFace graphical user interface toolkits. These frameworks are not widely used outside of the Eclipse ecosystem. As we have noted, a number of participants in our field study indicated that they hoped that our tool would be available for other IDEs and other programming languages. Indeed, many concepts (including regular expressions, colors and database queries) are similar across languages. To better support IDE-independence and language-independence, we chose instead to build palettes using HTML and Javascript. Javascript was among the most well-known languages in our study, just behind Java, C and C++, and is highly flexible. A number of user interface libraries are available, such as jQuery\footnote{http://jquery.com/} (which we used to implement our palettes). All major browsers now feature sophisticated debuggers and run-time inspection facilities that ease debugging and testing. Deploying palettes using standard URLs is also simpler than attempting to integrate their source code directly into Java libraries and packages and eases the process of incremental development and deployment. All major windowing toolkits feature a web browser control that can allow for the display of these palettes within the editor environment. In Eclipse, we used the standard SWT Browser control.

\subsection{Palette API}
In order to allow palettes to communicate with the surrounding editor environment, a simple Javascript interface was injected. These functions were implemented by the editor plug-in, but appear as ``native'' browser functions to Javascript code. The API consists of the following methods, accessed through a global object named \verb|graphite|:
\begin{itemize}
	\item \verb|insert(str)|

	Inserts the specified string at the cursor and closes the palette. The appropriate amount of indentation is automatically inserted after any newlines in this string.

	\item \verb|cancel()|

	Closes the palette without inserting a string.

	\item \verb|getSelectedText()|

	Returns the text that is currently selected by the user in the editor, or an empty string if no text has been selected. This method is used to implement reinvocation.

	\item \verb|getIDE()|

	Returns a string that specifies the IDE that is being used, ``Eclipse'' in our implementation.

	\item \verb|getLanguage()|

	Returns a string that specifies the programming language that is being used, ``java'' in our implementation.
\end{itemize}

By limiting the complexity of this API, we reasoned that developers would be able to create specialized palettes more easily. An additional benefit is that the developers of other editing environments are also able to create plug-ins that support Graphite palettes with minimal effort.

To access this API, the palette must include a small script named \verb|graphite.js| into their page. This file interacts with the host editor to create the graphite object and provides an alternative implementation of these methods when the palette is being tested in a web browser.

\subsection{Palette Discovery}
\begin{figure*}\begin{center}
\includegraphics[scale=.6]{regex.png}\end{center}
\caption{The regular expression palette developed using Graphite and used by subjects in the treatment group of the user study described in Sec. VI.}
\end{figure*}

Graphite currently provides two methods for associating a palette with a particular type so that the Graphite plug-in is able to discover it when the code completion menu is invoked by a developer.

\subsubsection{Annotation-based}
For user-defined classes that the palette author has the authority to modify, the \texttt{@GraphitePalette} annotation associates a palette with the class. The annotation must specify the URL of the palette, and can contain some other optional information (e.g. the description that is shown in the code completion menu). This  allows library providers to provide palettes that are specialized to their libraries and distribute them alongside their code. In this way, the users of the libraries are not required to discover that an external tool exists. If the user does not have the Graphite plug-in for Eclipse installed, the annotation has no effect.

\subsubsection{Explicit}
In cases where the palette developer cannot modify a class directly (such as palettes for classes in the Java standard library), a second method exists to explicitly associate a palette with the fully qualified name of a class. Currently, this feature is exposed as a preference pane in the Eclipse IDE. To facilitate automated installs in the future, a preferences file may also be added to the system.

\subsection{Design Trade-Offs}
The design that we have described is light-weight, highly flexible and highly responsive. It lays foundations for IDE and language portability. However, several trade-offs have also been made to achieve these benefits.

\begin{itemize}
\item Because palettes are implemented in Javascript, any differences between Java and Javascript become evident. For example, Javascript color names are, in two cases, slightly different from the corresponding Java color name. The Javascript regular expression engine is also slightly different. Although a Java applet could be embedded to provide necessary interoperability in cases where these differences are critical, this is more difficult than it would be if the palettes were implemented using the same language as the code.
\item The palette user interface must stay within its bounding box -- user interface elements like pop-up menus (unless they are provided by the browser itself) are thus more difficult to implement and may require additional API support in the future.
\item Several use cases could benefit from greater access to the surrounding code, or even the surrounding project. Implementing this modularly is particularly difficult, and the Eclipse API does not easily allow for  serialization of the sum of its contextual knowledge for consumption by a palette.
\item Implementing a reinvocation mechanism that is robust to changes in the code and does not require the user to select the relevant information is difficult because this would require that the palette  parse all the code to find relevant information. This also would require richer insertion APIs that allow for  ``dead'' code to be removed. This puts a greater burden on palette developers and errors could affect user's code in subtle ways. The small inconvenience of having to select code before reinvoking a palette may be balanced by the knowledge that no other portions of the code will be modified inadvertently by that palette.
\end{itemize}

\subsection{Palettes}
We implemented two full-featured palettes to demonstrate and test the functionality of this framework: a regular expression palette (Figure 5) and a color selection palette (Figure 1).

\subsubsection{Regular Expressions}
Writing correct regular expression patterns is difficult. The regular expression palette, associated with the \verb|Pattern| class in the Java standard library, allows users to enter and test regular expression patterns interactively before inserting them into their code.  The focal point of this palette is the pattern input area. As the user enters a pattern into this input area, syntax errors are indicated with a red background; the background remains white when the pattern is valid.

In addition to the pattern, a regular expression consists of flags that can change its matching behavior in various ways. Our palette allows users to toggle the case-sensitivity flag of the regular expression using a small checkbox labeled \verb|Ignore Case|, placed next to the input area. A keyboard shortcut is also available, indicated using the underlined letter (Ctrl+I).

To allow users to test the behavior of the regular expression that they have entered, two columns are available below the expression input area. The left column contains an input area labeled ``Should match...'' and the right column contains an input area labeled ``Should NOT match...''. Users use these to enter lists of test  strings into each column. The background colors behind these strings change to indicate whether the regular expression that has been entered matches or does not match that string. Green indicates that the pattern  {\it matched} the string and red indicates that it does {\it not} match, regardless of the column that the test is in (this scheme was chosen based on feedback from an initial pilot study of this palette.) A key describing this color scheme is displayed after the first test has been entered (not shown).

Users navigate between the three input areas using standard \verb|Tab| cycling behavior. The label in each text area remains visible until some input has been entered, rather than disappearing immediately on focus. When a user is satisfied with the regular expression that they have entered, she can press the \verb|Enter| key to insert the appropriate Java source code. Because Java requires that the regular expression pattern be placed inside a string literal, additional escape sequences are needed in front of backslashes. This can be tedious and error-prone if done manually. Because the palette is responsible for generating this literal, however, this is no longer a problem. In addition to the source code itself, the tests are retained in a comment beginning on the next line. The user can leave these as documentation or use them to create more formal unit tests within their preferred testing framework.

If the user wishes to modify the regular expression or change the test set, she can highlight the code that was inserted and then invoke the palette once again. The palette parses the selected text to extract the regular expression and tests. Although this is slightly inconvenient in that it requires that the user select the relevant portion of the code, it is the most robust method for implementing this feature -- it can easily handle changes to the regular expression or tests and it does not require additional logic in the palette to determine which portions of the code are relevant, and should subsequently be replaced if a new regular expression is entered.

\subsubsection{Color Selection}
We omit a description of the full behavior of the Color selection palette due to space limitations. 
%\section{Future work}
%
%\subsection{Graphite plug-in for different IDEs}
%As all the palettes are independent to Eclipse, We speculate that implementing Graphite plug-in for other IDEs will not be too difficult. Implementing a web-based aiding tool and sharing that tool among different IDEs with minimal effort will be a good idea.
%
%\subsection{Improving the Graphite API for Javascript}
%The current version of Graphite API for Javascript is not powerful enough. For example, the palettes cannot add comments before current line, nor can it delete or replace code fragment which was not highlighted by the user. By providing more effective and powerful API, the palette writers can write the palettes more quickly and make more useful palettes.
%
%\subsection{More palette examples}
%Due to the limited time, we could only provide two example palettes so far. More palette examples for various types might inspire the library providers and palette writers to design better palettes.
%
%\subsection{Keyword invocation model / online palette repository}
%Although Keyword invocation has many expected benefits, it is currently not supported. This model will make Graphite language-independent, and improve discoverability of the palettes by allowing users to search the palette with keywords. Palette search will work better if a shared online palette repository is built and maintained.
%
%

% An example of a floating figure using the graphicx package.
% Note that \label must occur AFTER (or within) \caption.
% For figures, \caption should occur after the \includegraphics.
% Note that IEEEtran v1.7 and later has special internal code that
% is designed to preserve the operation of \label within \caption
% even when the captionsoff option is in effect. However, because
% of issues like this, it may be the safest practice to put all your
% \label just after \caption rather than within \caption{}.
%
% Reminder: the "draftcls" or "draftclsnofoot", not "draft", class
% option should be used if it is desired that the figures are to be
% displayed while in draft mode.
%
%\begin{figure}[!t]
%\centering
%\includegraphics[width=2.5in]{myfigure}
% where an .eps filename suffix will be assumed under latex, 
% and a .pdf suffix will be assumed for pdflatex; or what has been declared
% via \DeclareGraphicsExtensions.
%\caption{Simulation Results}
%\label{fig_sim}
%\end{figure}

% Note that IEEE typically puts floats only at the top, even when this
% results in a large percentage of a column being occupied by floats.

% An example of a double column floating figure using two subfigures.
% (The subfig.sty package must be loaded for this to work.)
% The subfigure \label commands are set within each subfloat command, the
% \label for the overall figure must come after \caption.
% \hfil must be used as a separator to get equal spacing.
% The subfigure.sty package works much the same way, except \subfigure is
% used instead of \subfloat.
%
%\begin{figure*}[!t]
%\centerline{\subfloat[Case I]\includegraphics[width=2.5in]{subfigcase1}%
%\label{fig_first_case}}
%\hfil
%\subfloat[Case II]{\includegraphics[width=2.5in]{subfigcase2}%
%\label{fig_second_case}}}
%\caption{Simulation results}
%\label{fig_sim}
%\end{figure*}
%
% Note that often IEEE papers with subfigures do not employ subfigure
% captions (using the optional argument to \subfloat), but instead will
% reference/describe all of them (a), (b), etc., within the main caption.

% An example of a floating table. Note that, for IEEE style tables, the 
% \caption command should come BEFORE the table. Table text will default to
% \footnotesize as IEEE normally uses this smaller font for tables.
% The \label must come after \caption as always.
%
%\begin{table}[!t]
%% increase table row spacing, adjust to taste
%\renewcommand{\arraystretch}{1.3}
% if using array.sty, it might be a good idea to tweak the value of
% \extrarowheight as needed to properly center the text within the cells
%\caption{An Example of a Table}
%\label{table_example}
%\centering
%% Some packages, such as MDW tools, offer better commands for making tables
%% than the plain LaTeX2e tabular which is used here.
%\begin{tabular}{|c||c|}
%\hline
%One & Two\\
%\hline
%Three & Four\\
%\hline
%\end{tabular}
%\end{table}

% Note that IEEE does not put floats in the very first column - or typically
% anywhere on the first page for that matter. Also, in-text middle ("here")
% positioning is not used. Most IEEE journals/conferences use top floats
% exclusively. Note that, LaTeX2e, unlike IEEE journals/conferences, places
% footnotes above bottom floats. This can be corrected via the \fnbelowfloat
% command of the stfloats package.

\section{User Study Design}

We conducted a comparative study to evaluate the usefulness and usability of a Graphite palette in a constrained setting -- writing regular expressions. We used a between-subjects design by randomly splitting the participants into two groups: a control group, where subjects were not shown nor able to use the palette, and a treatment group, where subjects were shown a different palette and allowed to discover and use the regular expression palette if they wished.

\subsection{Process}
\subsubsection{Preliminary Survey}
The participants were first asked to fill out the pre-survey form about their prior experiences with various programming languages, integrated development environments (IDEs), and regular expressions, based on the previous survey that we had conducted. Of note, the subjects in this study were considerably less skilled with Java and regular expressions than in the previous survey, likely due to a sampling effect -- our subjects were all PhD students at Carnegie Mellon.

\subsubsection{Training}
Only the participants in the treatment group were shown how to invoke Graphite palettes in the context of an  Eclipse code editor with a palette for the Color class. We chose to demonstrate the tool using a color palette instead of the regular expression palette itself because we wanted to simulate the condition where a user had discovered the palette naturally. The demonstration of the Color palette was brief, taking about two minutes.
We then described the nature of the task to the subject and allowed them to begin, giving them 45 minutes to complete all tasks, in any order.

\subsection{Within vs. between-subjects design}

We decided to use a \textit{between-subjects design} because a within-subjects design would have required us to produce pairs of tasks with equal difficulties. This turned out to be very difficult. Second, we could not easily ignore a learning effect during the experiment if we had used a within-subjects design. We observed that this effect was quite strong, as most subjects hadn't used regular expressions recently.

\subsection{Participants}

We recruited 7 subjects from Computer Science Department and Institute for Software Research at Carnegie Mellon University. All the 7 subjects were PhD students. The subjects were randomly assigned into two groups: 4 subjects were assigned to the control group, and the other 3 subjects to the treatment group. There were six male participants and one female participant. Participants were compensated in the amount of 15 dollars for their participation.

\subsection{Tasks}

There were a total of 9 tasks to be completed in 45 minutes. The first 6 tasks involved writing regular expressions to validate various data formats (e.g. temperatures), and the remaining 3 tasks involved writing regular expressions to retrieve data from a document. The participants were allowed to move back and forth while doing the experiment, and they were allowed to use any external resources, including the internet, local console, and so forth (we did not observe any usage of programs other than the web browser, however). The only restriction we placed on their activity was that they were NOT allowed to directly search for the answer to a task online online. We omit descriptions of each individual question due to space limitations.

%
%\begin{itemize}
%	\item Data format validation tasks (total 6 tasks)
%	
%	\begin{enumerate}
%		\item US phone numbers
%		
%		For simplicity, it was assumed that any digit can be used in any place in the phone number, even though there is a notion of valid range for the area code.
%		
%		\item Times
%		
%		Hours and minutes only.
%		
%		\item Temperatures
%		
%		Only Fahrenheit and Celcius were used, and subjects were instructed to assume the degree symbol had not been used.
%		
%		\item Dates
%		
%		We did not place any restrictions on the format for dates. This task is quite difficult to solve generally. There are many valid formats for expressing dates, and the valid range of the day part depends on the specific year and month. Many participants were stuck on this task and chose to skip it and came back later.
%		
%		\item IPv4 addresses
%		
%		\item URLs starting with \texttt{http://} or \texttt{https://}
%		
%	\end{enumerate}
%	
%	\item Data searching tasks (total 3 tasks)
%	
%	\begin{enumerate}
%	\setcounter{enumi}{6}
%		\item Finding a fields declaration (member variable) in a Java class
%		
%		It was assumed that all the field declarations start with one of the access modifier keywords: \verb|public|, \verb|protected|, or \verb|private|.
%		
%		\item Finding a string literal in a Java file
%		
%		\item Finding an \texttt{img} tag in an HTML file
%		
%	\end{enumerate}
%	
%\end{itemize}
%
\subsubsection{Task Structure}

The subjects were given a skeleton class which contained the instruction for how to complete the task and a method stub called \texttt{getRegexPattern} for each task. They are asked to fill the body of the \texttt{getRegexPattern} method. The \texttt{Task1.java} - \texttt{Task9.java} files were listed in the Eclipse project explorer so that the subjects could easily open and reopen the tasks in any order that they wanted.

%\begin{figure}[h]%
%\includegraphics[width=\columnwidth]{Task1.png}%
%\caption{The initial content of \texttt{Task1.java} file}%
%\label{task1}%
%\end{figure}

\section{User Study Results}
After reading the prompt for the first task, all subjects began by searching for and reviewing documentation related to regular expressions. In all but one of these cases, the subject looked at the API documentation for the Java Pattern class during this initial review. The remaining subject, a member of the control group, referred to quick reference documentation provided by an external tool. In all cases, the documentation was left accessible in a browser window throughout the study, often displayed simultaneously alongside the coding window due to the large screen available for the study.

\subsection{Control}
In our previous online survey, we had found that no single strategy for writing regular expressions dominated the others. About 40\% of participants indicated that they would use an external tool, 30\% indicated a preference for writing a separate test script, and about 13\% of participants chose the guess-and-check strategy. Although our sample size is too small to make a quantitative comparison, we did observe each of these strategies in the control group.

One subject began by using an external tool called regexpal.com. After writing a full regular expression and attempting to test it using the tool, the subject became dissatisfied with it and switched to a different tool, regextester.com. This tool too appeared to be unsatisfactory, as all subsequent tasks were performed without the aid of external tools or tests of any kind.

The other subjects chose to write their regular expressions directly within Eclipse from the beginning. One of these subjects never compiled or checked the accuracy of the regular expressions beyond making sure Eclipse errors were addressed, while the other two attempted to write Java stubs within the test files to test the regular expression that they had written. They experienced considerable difficulty with this task, with one subject taking over 10 minutes to write the test code. The code could only check one example at a time and the subject used it to check a single positive example per task.

In designing the regular expression palette, we had hypothesized that users would experience difficulties with two particular aspects of the Java Pattern API: that it used a factory pattern for instantiation, as opposed to the standard instantiation construct (new), and that it required special care with escape sequences -- escape sequences in the pattern must themselves be escaped, because the regular expression is written as a Java string literal. We observed difficulties with both of these issues in the control group.

One of the four control subjects tried the \verb|new Pattern| construct. An Eclipse error alerted the subject to the problem shortly thereafter. After referring to an example in the inline API documentation for the class, the subject was able to correct this error. The remaining subjects all noticed this example beforehand, and were able to avoid this problem.

Three of the four control subjects experienced significant difficulties associated with escape sequences. One subject recognized the error fairly quickly after Eclipse complained that the escape sequences used in the pattern were invalid (they were, in fact, valid escape sequences for regular expression patterns, but not for string literals.) The subject continued to miss escape sequences occasionally throughout the remainder of the study, but was able to fix them quickly after Eclipse alerted the subject to the error. For the other two subjects, the problem was more severe. In both cases, they noticed the error that Eclipse gave, but thought it indicated a problem with their pattern itself. One subject thought that the uses of symbolic escape sequences like \verb|\(| were incorrect. He decided to replace these with ASCII escape sequences (\verb|\050|) after looking up an ASCII conversion table. This was an unnecessary and overly complex solution to the problem. The other subject thought that the problem was in his use of the whitespace escape sequence, \verb|\s|. Thus, he  spent several minutes looking up how to match whitespace correctly. In doing so, he found a description of the double-escape problem in Java and was able to fix the problem after that.
\subsection{Treatment}
One subject was already familiar with an external tool, regexpal.com (also used by a subject in the control group), and used it variously throughout the task. In most cases, the subject used the external tool's quick reference documentation while using our palette for authoring and testing. At other times, the subject used the external tool itself. The subject indicated that in some cases, he was not sure that our palette was free of bugs (when something unexpectedly matched or did not match\footnote{We did not observe any actual errors related to regex matching in our tool.}), and also because the external tool provided syntax and substring highlighting, a useful feature for complex regular expressions. When done with the external tool, the subject pasted the pattern into our tool to generate the appropriate code. As such, the subject had no difficulties with escaping or factory pattern instantiation.

The other two subjects in the treatment group did not use any external tools. Both subjects decided not to use our palette for the first task (likely because they forgot that it was available, since they spent some time looking up API documentation after our initial demo with the color palette.) They both recognized the need for factory pattern instantiation and also both had an initial error related to escaping that they were able to resolve before moving on. One of these subjects also began to write tests within a \verb|main| method.

Beginning with the second task, both subjects remembered that a palette may be available and were able to invoke it correctly. They all recognized that the primary input box was where the regular expression should be entered and that the two other input boxes were for positive and negative tests, respectively. Two of the subjects did not initially realize that multiple tests could be entered, and that entering even a single test required pressing the \verb|Enter| key. One subject never fully understood this, relating after the study that he thought that the palette would notify him when he was done if the example that he had entered was not matched by the pattern. The other subject was able to use the test case input boxes correctly after initial experimentation. This was perhaps due to the wording of the prompts under these entry boxes: ``enter \textbf{a} [positive \verb1|1 negative] test case above''. This problem was corrected in a version of the palette developed following this study.

Two of the subjects used the reinvocation feature of the palette, but neither highlighted the test cases, meaning that they had to enter new test cases every time they reinvoked the palette. This may have been due to the fact that our example with the Color palette involved only a single line of text (unlike the example in Figure 1). The third subject never used this feature.

Two of the subjects expressed confusion about the meaning of the green and red backgrounds on the test cases. Although we provided a key saying that green meant that ``pattern matches string'', the headings of the two columns: ``[should \verb1|1 should NOT] match:'', may have caused confusion. We had changed the meaning of these colors due to feedback from our initial presentation of the tool in class, so it is clear that regardless of the interpretation given, some subjects were confused.

Despite these difficulties, however, the basic functionality of the palette seemed to help all of the subjects. None struggled with issues related to escaping and instantiation, for example. Although the sample size is not large enough to make conclusive or quantitative judgements, it was observed that the treatment group completed more of the tasks than the test group on average (7 tasks for the treatment group vs. 6 for the control group). All of the subjects in the treatment group, as well as members of the control group who were shown the palette after the study, indicated that they felt that the palette was helpful, and several provided specific suggestions for improvements.
%\section{Future Work}
%
%\subsection{Palette Design Improvements}
%The observations above suggest several potential improvements for the regular expression palette. Some of the problems appeared to be due to the subject's lack of familiarity with the Graphite system. In particular, subjects did not highlight the full text needed to reinvoke the palette, likely due to the the limited training protocol. By showing only a single, simple example, the subjects were not able to generalize the reversibility mechanism to multiline completions. If the training period had included such an example, we suspect that this would not have been a problem. 
%
%In other cases, more explicit labels in the palette may have helped. The subjects began by entering only a single example because the label on the palette used the singular article, ``a''. We suspect that we could have avoided this problem if the label was more explicit: ``Create positive examples by entering them above and pressing ENTER''. Confusion over the meaning of the green and red backgrounds may also have been avoided by using a clearer label for the key, accompanied by a label explaining the meaning of each column more explicitly.
%
%In conjunction with these improvements, it would be useful to have an explicit ``Help'' button that allowed users to view documentation or a training video for the palette. On first invocation, it may also have been helpful to invoke this video automatically (though this would require tool support for associating data with the user.)
%
%Several subjects suggested new features for the regular expression palette, including syntax highlighting, substring highlighting and an explicit option to match the whole string, rather than any part of the string (a problem encountered by several subjects in both groups). They also suggested including inline quick reference documentation and regex snippets directly within the palette. Other people (not a part of the study) have suggested other features, including a tool for building regular expressions directly from examples, and a tool for generating examples automatically from regular expressions. Examples of such tools exist online.
%
%\subsection{Study Design Improvements}
%
%Some of the limitations in our study design, combined with the small size of our subject pool, made any sort of quantitative analysis difficult. In this section, we suggest some possible improvements.
%
%	\paragraph{Task Difficulty} Subjects were sometimes ``stumped'' by some of the more difficult tasks. It would perhaps be helpful to remove some of these excessively difficult tasks. For example, validating US phone numbers and dates turned out to be much more difficult than the others.
%	
%	\paragraph{Giving a time limit for each task} Because we only imposed a time limit for the whole experiment, it was difficult to compare performance on an individual task, in terms of time spent. However, imposing a per-task time limit may introduce concerns about the learning effect -- most subjects spent a significant amount of time looking up documentation in the early part of the study. One possibility may be to introduce dummy tasks at the beginning of the experiment.
%	
%	\paragraph{Using other types of Graphite palettes} This study does not evaluate the whole Graphite plug-in architecture, and many of the issues may have been due to the regular expression task and palette specifically. Designing a study to effectively evaluate the design as a whole would likely require building several more high-quality palettes.
%	
%	[Evaluating performance] It is difficult to measure accuracy of provided regular expressions due to their inherent complexity. We could not run an automated test suite on the subject responses because many subjects had syntax errors or other problems in their final submission that would have caused trivial failures. 	Poor-quality regular expressions written in a relatively short amount of time may also be hard to compare to high-quality regular expressions written in a longer amount of time. 
%		
\section{Related Work}
In addition to the references provided in the introduction, some other concepts are related to active code completion.

\subsection{Active Libraries}
We named this technique {\it active code completion} because of its relation to the general concept of {\it active libraries} \cite{activelibraries}. Active libraries are libraries that contain program logic that is invoked at either compile-time or, here, design-time.

\subsection{Visual Languages}
Because active code completion involves graphical user interface elements but ultimately generates textual source code representations, it can be considered a hybrid approach that borrows interaction techniques from visual languages while remaining compatible with conventional programming languages. This hybrid approach may help address some of the usability challenges previously associated with visual languages (cf. \cite{oai:CiteSeerXPSU:10.1.1.110.3146}).

Editing environments like Barista \cite{Ko06} has previously been developed that merge concepts from both conventional text based and structured editors. Barista allows developers to use alternative code representations. It differs from Graphite in that these representations are built directly into the editor (although adding new representations is possible), and that Barista requires the users to use its own model-view-controller based language, which is likely to be unfamiliar to many users. In future work, we hope to explore an extensible, keyboard-driven, code-generation based approach that leverages structured code representations to eliminate the difficulties associated with reinvocation and maintaining palette state described in this paper.

\subsection{Specific IDE Features}
There exist some IDE features that have been specifically designed for certain types. For example, CodeRush has a ``show color dialog'' \cite{CodeRush} which allows the developers to launch a color picker directly from the code editor. Resharper \cite{Resharper} also has a color assistance tool which previews the actual colors right beside the code completion menu items. However, these IDE specific features are hard-coded -- user-defined types cannot provide similar functionality. Recent versions of Visual Studio have started to explore user-defined palettes for widgets in the windowing system.

\section{Conclusion}
Motivated by evidence that integrating highly-specialized tools directly into a developer's workflow is useful, we have developed the concept of active code completion as a generalization of conventional code completion. We validated the usefulness and generated a number of use cases and design constraints for this concept with an extensive survey of professional developers. Based on these findings, we developed an active code completion architecture that makes several novel design decisions to ease development, deployment and discovery of user-defined palettes. We created a palette for regular expression tasks and validated its usefulness with a user study. We claim that this provides evidence for the more general claim that integrating palettes into a code editor is useful, and that active code completion systems like Graphite considerably ease this process.

% conference papers do not normally have an appendix

% use section* for acknowledgement
\section*{Acknowledgment}

% trigger a \newpage just before the given reference
% number - used to balance the columns on the last page
% adjust value as needed - may need to be readjusted if
% the document is modified later
%\IEEEtriggeratref{8}
% The "triggered" command can be changed if desired:
%\IEEEtriggercmd{\enlargethispage{-5in}}

% references section

% can use a bibliography generated by BibTeX as a .bbl file
% BibTeX documentation can be easily obtained at:
% http://www.ctan.org/tex-archive/biblio/bibtex/contrib/doc/
% The IEEEtran BibTeX style support page is at:
% http://www.michaelshell.org/tex/ieeetran/bibtex/
%\bibliographystyle{IEEEtran}
% argument is your BibTeX string definitions and bibliography database(s)
%\bibliography{IEEEabrv,../bib/paper}
%
% <OR> manually copy in the resultant .bbl file
% set second argument of \begin to the number of references
% (used to reserve space for the reference number labels box)
\bibliographystyle{IEEEtran}
%\bibliography{graphite-icse12}
\bibliography{../research}

% that's all folks
\end{document}

### 2012-03-18 vlhcc11/survey-part.tex
Initial paper move.

--- 2012-03-18 icse12/graphite-icse12.tex
@@
 %
-\documentclass[10pt, conference, compsocconf]{IEEEtran}
-% Add the compsocconf option for Computer Society conferences.
+\documentclass[conference]{IEEEtran}
+% Add the compsoc option for Computer Society conferences.
 %
@@
 % \documentclass[conference]{../sty/IEEEtran}
-
-\usepackage{rotating}
 
@@
 % The documentation is contained in the cite.sty file itself.
+
+\usepackage{enumerate}
+\usepackage{rotating}
+\usepackage{url}
 
@@
 
-\usepackage{alltt}
-\usepackage{hyperref}
-\hypersetup{pdfborder={0 0 0 0}}
-
 \begin{document}
@@
 % author names and affiliations
-% use a multiple column layout for up to two different
+% use a multiple column layout for up to three different
 % affiliations
-
 \author{\IEEEauthorblockN{Cyrus Omar, YoungSeok Yoon, Thomas D. LaToza, Brad A. Myers}
-\IEEEauthorblockA{Carnegie Mellon University, 
+\IEEEauthorblockA{School of Computer Science\\
+Carnegie Mellon University\\
 Pittsburgh, PA, USA\\
-\small{\texttt{\{comar,youngseok,tlatoza,bam\}$@$cs.cmu.edu}}\\
-\small{\texttt{\url{http://www.cs.cmu.edu/~NatProg/graphite.html}}}}
+\{comar,youngseok,tlatoza,bam\}@cs.cmu.edu}
+\texttt{\url{https://github.com/cyrus-/graphite}}
 }
@@
 % make the title area
-\maketitle
-
-\begin{abstract}
-Code completion menus have replaced standalone API browsers for most developers because they are more tightly  integrated into the development workflow.
-Refinements to the code completion menu that incorporate additional sources of information have similarly been shown to be valuable, even relative to standalone counterparts offering similar functionality.
-In this paper, we describe {\em active code completion}, a technique that enables the integration of a much broader array of interactive and highly-specialized code generation interfaces, called {\em palettes}, directly into the editor environment.
-Using several empirical methods, we examine the contexts in which such a system could be useful, describe the design constraints governing the system architecture as well as particular code completion interfaces, and design one such system, named Graphite, for the Eclipse Java development environment.
-Using Graphite, we implement a palette for writing regular expressions as our primary  example and conduct a controlled user study that provides further evidence in support of the claim that integrating specialized tools directly into the editor environment is both feasible, particularly if aided by an  active code completion system like Graphite, and valuable to professional developers.
-\end{abstract}
-
-\begin{IEEEkeywords}
-code completion, development environments
-
-\end{IEEEkeywords}
+% \maketitle
+
+% IEEEtran.cls defaults to using nonbold math in the Abstract.
+% This preserves the distinction between vectors and scalars. However,
+% if the conference you are submitting to favors bold math in the abstract,
+% then you can use LaTeX's standard command \boldmath at the very start
+% of the abstract to achieve this. Many IEEE journals/conferences frown on
+% math in the abstract anyway.
+
+% no keywords
 
@@
 % creates the second title. It will be ignored for other modes.
-\IEEEpeerreviewmaketitle
-\section{Introduction}
-Software developers today make heavy use of the code completion support found in modern source code editors  \cite{murphy_how_2006}.
-Most editors provide code completion in the form of a floating menu containing  contextually-relevant variables, fields, methods, types and other code snippets.
-By navigating and selecting from this menu, developers are able to avoid many common spelling and logic errors, eliminate unnecessary keystrokes and explore unfamiliar APIs without incurring the mental overhead associated with switching to an external documentation tool or API browser.
-% The way code completion menus brought documentation systems into the editor, active code completion interfaces bring external tools into the editor.
-
-Several refinements and additions to the code completion menu have previously been suggested in the literature.
-These have focused on leveraging additional sources of information, such as databases of usage history \cite{robbes_how_2008}, examples extracted from code repositories \cite{bruch_learning_2009} and crowdsourced information \cite{mooty_calcite:_2010}, to increase the relevance and sophistication of the featured menu items.
-As with the standard form of code completion, many of these sources of data can also be utilized via external tools (e.g.
-Calcite \cite{mooty_calcite:_2010} uses information that could already be accessed using the Jadeite \cite{conf/vl/StylosFYM09} tool).
-Empirical evidence presented in these studies, however, indicates that directly integrating these kinds of tools directly into the editor environment can be more effective.
-
-In all of these systems, the code completion interface has remained primarily menu-based.
-When an item is selected, code is inserted immediately, without further input from the developer.
-These systems are also limited in that they are difficult to modularly extend.
-A fixed strategy determines the completions that are available, leaving library providers unable to specify new domain-specific or contextually-relevant items.
-In this paper we propose a technique called {\it active code completion} that eliminates these restrictions.
-This  makes developing and integrating a broad array of highly-specialized developer tools directly into the editor, via the familiar code completion command, significantly simpler.
-This technique is motivated by the evidence discussed above and further evidence provided in this paper that developers prefer, and make more effective use of, tools that do not require leaving the immediate editing environment.
-
-In this paper, we discuss active code completion in the context of object construction. For example, consider the following Java code fragment:
-
-\begin{alltt}
-  public Color getDefaultColor() \{
-      return \textvisiblespace
-\end{alltt}
-
-\begin{figure*}\label{color}
-\begin{center}
-\includegraphics[scale=0.8]{color_palette.png}\end{center}
-\caption{(a) An example code completion palette associated with the \texttt{Color} class. (b) The source code generated by this palette.}
-\end{figure*}
-
-If the developer invokes the code completion command at the indicated cursor position, the editor  looks  for a {\it palette definition} associated with the {\it type} of the expression being entered, which in this case is  \verb|Color|.
-If an associated palette is found, a menu item briefly describing this palette is added to the standard code completion menu.
-When selected, the corresponding palette is shown, replacing the standard code completion menu.
-Figure 1a gives an example of a simple palette that may be associated with the \verb|Color| class.
-
-The developer can interact with such palettes to provide parameters and other information related to her intent, and receive immediate feedback about the effect these choices will have on the behavior of the object being constructed.
-When this interaction is complete, the palette is responsible for generating appropriate source code for insertion at the cursor.
-Figure 1b shows the code that is inserted after the user presses the \textsc{enter} key.
-
-In accordance with rational design practices, we sought to address the following questions before designing and implementing our active code completion system:
+% \IEEEpeerreviewmaketitle
+
+\section{Developer Survey}
+
+Prior to implementation, we conducted a large online survey of professional software developers in order to extract criteria that govern whether our system would be useful to them.
+More specifically, we sought to address the following questions:
 
 \begin{itemize}
-\item What {\it specific} use cases exist for this form of active code completion in a professional development setting? 
-\item What {\it general} criteria are common to types that would and would not benefit from an associated palette?
-\item What are some relevant usability and design criteria for palettes designed to address such use cases?
-\item What capabilities must the underlying active code completion system provide to enable these use cases and user interface designs?
+\item What are specific use cases for active code completion in a professional setting? 
+\item What are some functional criteria that predict whether a class would benefit from active code completion?
+\item What are some usability criteria that should govern palette interface designs?
+\item Which capabilities must a active code completion architecture support to enable these use cases and designs?
 \end{itemize}
 
-To help us answer these questions, we conducted a large online survey of professional developers (Section 2).
-Their responses, along with information gathered from informal interviews and code corpus analyses, revealed a number of interesting use cases (Section 3) and non-trivial functional requirements for palette interfaces as well as the underlying active code completion architecture (Section 4).
-Notably, many of these findings can be applied to development tools for professional developers more generally, not simply those integrated into the editor in the way we have shown.
-
-Next, we describe Graphite, an Eclipse plug-in that provides active code completion for users of the Java programming language (Section 5).
-We describe several interesting design choices that were made to satisfy many of the requirements discovered in our preliminary investigations and briefly examine the trade-offs that were necessary due to Java's purely textual program representations and Eclipse's limited extension APIs.
-
-Finally, we conduct a controlled lab study with a more complex palette, implemented using Graphite, that assists  developers as they write regular expressions (Section 6).
-The data and observations gathered from this study provide  specific evidence in support of the broader claim that highly-specialized tools that are integrated directly with the editing environment are particularly useful, and that an active code completion system like Graphite makes developing, deploying and discovering such tools fundamentally simpler.
-
-\section{Survey Results}
-To validate our general conceptualization of active code completion, develop concrete design criteria and create a concrete list of use cases, we began by conducting a large survey of professional software developers.
-
 % In order to convey the active code completion concept clearly, we came up with three different use cases where we belive the active code completion technique will benefit developers frequently.
-For each use case, we designed three mockup screenshots and showed them in sequence: before invoking the palette, during the interaction with the palette, and after the palette is closed and code is generated.
+For each use case, we designed three mock-up screenshots and showed them in sequence: before invoking the palette, during the interaction with the palette, and after the palette is closed and code is generated.
 By doing this, the respondents could understand the concept precisely and give useful feedback relevant to the above questions.
-\subsection{Survey Methods}
-We recruit participants for this survey\footnote{https://www.surveymonkey.com/s/2GLZP8V} primarily from a popular programming-related discussion forum hosted on the popular website reddit.com\footnote{http://www.reddit.com/r/programming}.
-An additional 22 participants were recruited from the local computer science graduate student body.
-
-Recruitment materials in both cases stated that we were seeking developers ``familiar with an object-oriented programming language like Java, C\# or Visual Basic and an integrated development environment like Eclipse or Visual Studio''.
-Participants were told that the survey would take approximately 20 minutes to complete, and no reward was offered.
-Of the 696 people who started the survey, 473 participants (68\%) completed it.
-We examine the responses from completed surveys only in the analyses below.
+
+We posted a link to our survey on a popular  collaborative filtering website \cite{mehta_robust_2007} called {\tt /r/programming} on reddit.com\footnote{\texttt{http://www.reddit.com/r/programming}}.
+At the time of our study, the website listed approximately 340,000, though members were not required to registered to read the content,  and contained content that could be considered relevant to our intended audience.
+An additional 22 participants were recruited using a mass email to local computer science graduate students.
+The recruitment materials in both cases stated that we were seeking developers ``familiar with an object-oriented programming language like Java, C\# or Visual Basic and an integrated development environment like Eclipse or Visual Studio''.
+
+Participants were asked to take an online survey that would take approximately 20 minutes to complete.
+Of the 696 people who started the survey by answering the first question, 473 participants completed it.
+Of these, 16 participants were computer science graduate students at Carnegie Mellon and 457 were from \texttt{/r/programming}.
+
+Their responses revealed a number of interesting use cases and non-trivial design constraints for active code completion systems.
+Many of these findings can also be generalized to constrain the design of visual programming languages, particularly those that aim be useful to professional developers, rather than novice programmers (e.g., Barista \cite{ko_barista:_2006}).
+We summarize the survey questions and the results in the following subsections.
+% In fact, active code completion can be considered a hybrid approach that borrows interaction techniques from visual languages while remaining compatible with conventional representations of program logic.
+This may help address some of the usability challenges previously associated with visual languages (cf.
+[?]).
 
@@
 
-We first asked participants about their level of familiarity with several programming languages, on a five-point Likert scale ranging from ``None'' to ``Expert''.
-61.1\% of the participants indicated that they were an expert in at least one language, and an additional 35.7\% were ``very familiar'' with at least one language.
+We first asked their level of familiarity with several programming languages, on a five-point Likert scale ranging from ``None'' to ``Expert''.
+61.1\% of the respondents expressed themselves as experts on at least one language, and 35.7\% were very familiar with at least one language.
 On average, participants rated themselves as very familiar with Java, C, C++ and JavaScript, familiar with C\#, Python and PHP and somewhat familiar with Visual Basic and Perl.
 
-We also asked participants to select which integrated development environments (IDEs) and code editors that they were familiar with.
+We also asked participants to select the integrated development environments (IDEs) and code editors that they were familiar with.
 The Eclipse IDE was familiar to 87.1\% of participants.
 This was followed by Visual Studio at 66.0\%, Vi/Vim at 53.7\%, Netbeans at 37.7\%, Emacs at 24.8\% and IntelliJ IDEA at 16.4\%.
-Participants could also enter ``other'' choices and a number of editors and IDEs were entered, including Xcode, Textmate and Notepad++.
-
-\subsection{Palette Mockups}
-
-%\begin{figure}
-%  \includegraphics[width=\linewidth]{mockup-palette-small.png}
-%  \caption{An example mockup palette for regular expressions shown to the survey participants. In the survey, they were shown the surrounding Eclipse IDE together.}
-%  \label{regex}
-%\end{figure}
-%
-Next, we presented participants with a series of mockup palettes for a Color class (more complex than the one we ultimately implemented in Figure 1a), a regular expression class, and a SQL query class.
-Participants were also shown mockup screenshots demonstrating how a user could invoke the palette, and a mockup showing the code that would be inserted once a selection had been made.
-Before presenting each mockup, we gathered information about  the strategies that they would normally use to instantiate the class.
-
-For the Color class, the majority of participants indicated that they would look in the code completion menu (58.4\%) or in the class documentation (19.0\%) for a predefined constant if asked to instantiate an object corresponding to the color ``navy'' (which is not, in fact, a standard color in Java.) Another 14.0\% indicated that they would use an external tool (such as an image editor) to determine the RGB values corresponding to the color.
-
-Before asking participants about regular expressions and SQL queries, we asked participants to rate their familiarity with these concepts.
-Few participants (4\%) indicated that they were unfamiliar with regular expressions and no participants were unfamiliar with creating SQL queries, providing further evidence that our participants were not novice developers.
+Participants could also enter ``other'' choices.
+A number of editors and IDEs were entered, including Xcode, Textmate and Notepad++.
+None of these were dominant.
+
+\subsection{Mock-ups}
+
+\begin{figure}
+  \includegraphics[width=\linewidth]{mockup-palette-small.png}
+  \caption{An example mock-up palette for regular expressions shown to the survey participants. In the survey, they were shown the surrounding Eclipse IDE together.}
+  \label{regex}
+\end{figure}
+
+Next, we presented participants with a series of mock-up palettes for a Color class, a regular expression class (Fig.
+\ref{regex}), and a SQL query class.
+Participants were also shown a mock-up demonstrating how a user could invoke the palette, and a mock-up showing the code that would be inserted once a selection had been made.
+
+Before presenting each mock-up, we gathered information about the user's familiarity with the class and the strategies that they would likely use to instantiate the class under normal circumstances.
+For the Color class, the majority of participants indicated that they would look in the code completions menu (58.4\%) or in the class documentation (19.0\%) for a predefined constant.
+Another 14.0\% indicated that they would use an external tool to determine the RGB values.
+Few participants were unfamiliar with regular expressions and no participants were unfamiliar with SQL queries which provides further evidence that our participants were not novice developers.
 Fig.
-\ref{strategies} summarizes the strategies that participants generally preferred for instantiating regular expressions and SQL queries.
-Using an external tool was a common strategy in both cases, particularly for SQL queries, but several other strategies were also represented.
+\ref{strategies} summarizes their usual strategies for instantiating regular expressions and SQL queries.
 
@@
 \end{tabular}
-\caption{Distribution of responses to survey questions asking about typical strategies for writing regular expressions and SQL queries.}
+\caption{Typical strategies for regular expressions and SQL queries.}
 \label{strategies}
@@
 
-Finally, after showing the series of mockup screenshots, we asked participants to rate how useful the integrated palette would be to them if they needed to instantiate the corresponding class.
-The responses to this question for each palette are summarized in Fig.
+Participants were positive about our mock-ups.
+We asked the users to rate the usefulness of each mock-up palette, and their responses are summarized in Fig.
 \ref{frequencies}.
-In each case, more than half of the participants indicated that they would use active code completion at least some of the time.
-The regular expression palette was considered particularly useful while the color and SQL palettes showed a more reserved pattern of responses.
-
-In addition to asking for a simple rating for each palette, we also solicited open-ended comments.
-A large number of participants volunteered comments: 193 for the color palette, 129 for the regular expression palette and 142 for the SQL palette.
-These responses were highly valuable when developing the design criteria below and helped to explain the patterns observed in Fig.
-\ref{frequencies}.
-
-\section{Design Criteria}
-Using the information gathered from the survey as well as informal discussions with developers and researchers over the course of this study, we sought to develop design criteria constraining both the overall system design as well as the design of individual palettes.
-In the section headings below, the number of survey responses, summed over the three palette mockups, that contained the listed concern, as judged by the authors of this paper, are listed in parenthesis.
-Many of these concerns apply generally to various kinds of editor-integrated tools, not just those implemented using the active code completion mechanism that we focus on, so we hope these design criteria will be valuable to the developer tool design community more broadly.
-
-\subsection{Maintaining Separation of Concerns (183)}
-The most common issue developers had, particularly with the Color palette, was that it seemed to ignore the issue of separation of concerns by allowing developers to insert color constants directly into the program logic.
-Many developers noted that this is considered bad practice, or should be limited to the prototyping phase of a project.
-This concern was also expressed in responses to the SQL palette, which required inserting parameters to connect to a database so that the query could be tested.
-The code that was inserted included initialization steps needed to connect to the specific database that was entered, and several participants noted that much of this information should appear in an external resource file rather than inline with the program logic itself.
-Few participants made similar comments about the regular expression palette, however, indicating that regular expressions are considered a part of the program logic rather than data by most developers.
-
-This feedback suggests that tool and palette designers may wish to explicitly aprise users of relevant best practices and acknowledge that palettes that generate constant data may be most useful in the prototyping phase.
-It can be noted that when transitioning from a prototype to production-quality software the code generated by a palette or tool may be used as template that can be refactored to follow stricter policies as needed.
-We showed users the Color and SQL palettes in the context of a standard programming editor, but these comments indicate that resource file and  stylesheet editors may benefit from active code completion support as well.
-
-\subsection{Integration with Testing Frameworks (35)}
-The regular expression palette shown to the participants allowed users to immediately test a pattern against provided strings.
-These test strings and the results of performing the match were inserted as comments below the generated source code.
-A number of participants requested that explicit unit tests be generated instead.
-Although the palette interface conveyed the results of these tests immediately, participants remained adamant about explicit unit tests, likely due to concerns that future modifications to the generated code without reinvoking the palette might introduce bugs, or due to the desire to conform to standard testing practices.
-To support the generation of unit tests, however, the active code completion architecture must explicitly allow for code generation at locations other than at the cursor and handle the fact that statements generated after a code completion command are not necessarily encapsulated well enough to be immediately testable in isolation.
+For each palette, more than half of the respondents expressed that they would use it at least some of the time.
+Especially, people liked the regular expression palette a lot.
+
+\section{Design Implications}
+
+We analyzed the open-ended responses to extract design principles and constraints.
+We got 193 responses for the color palette, 129 for the regular expression palette, and 142 for the SQL palette.
+Also, we got total 119 suggestions about what other types of classes could benefit from similar types of palettes.
+The extracted principles and constraints are summarized in the following subsections.
+The parenthesized number following each design constraint indicates how many people commented on that issue.
+
+\subsection{System Design Constraints}
+
+System design constraints are the criteria that should be taken into account in the design of the overall system architecture.
+
+\subsubsection{Handling Separation of Concerns (183?)}
+
+Participants were very wary of including constant data into the source code directly.
+Many responses expressed general sentiment or specific preference for project-wide color theme or external resource file.
+Sometimes they explicitly suggested to add capability to handle the external resource files directly from the palette.
+Interestingly, most people seemed to consider regular expressions as part of the program logic.
+There were few people who complained about including the regular expression pattern string in the source code as opposed to the other two classes.
 
@@
 
-\subsection{Support for Reinvocation (19)}
-Several participants asked for the ability to reinvoke a palette from previously generated source code.
-In order to support this feature, the architecture must provide the palette with enough information to reconstruct its state.
-To complicate matters, however, users may wish to modify the generated code between invocations of the palette and have these modifications reflected in the palette's state upon reinvocation (e.g.
-modify the RGB values in the case of the \verb|Color| class).
-Moreover, there may be important aspects of the palette's state that are not directly available in the generated code, such as parameters controlling the palette's user interface.
-Indeed, associating tool-related metadata with code is known to be cumbersome in purely textual languages, since all metadata must be directly visible within comments or annotations.
-
-\subsection{Support for Palette Settings and History (41)}
-
-A related feature important to many participants is support for maintaining settings and usage history across invocations of the same palette at different code locations.
-For example, 20 participants requested that the Color palette include a list of recent or favorite colors, and 12 participants specifically inquired about whether the database connection information was maintained between invocations of the SQL palette.
-
+\subsubsection{Reversibility (19?)}
+
+Some participants expressed that it should be able to bring the palettes back when they want to modify the parameters after they used the palettes.
+
+\subsubsection{Palette Settings and State (41?)}
+
+Many participants wanted to be able to configure the palettes or have the palettes maintain state even after the palettes are hidden.
 % The followings are the comments that are related to this category.
@@
 
-\subsection{Support for Nested Expressions (13)}
-
-In all of the examples that we gave, the parameters entered into the palette interface corresponded to simple, constant expressions, rather than complex expressions referring to variables from the surrounding context.
-A number of participants noticed this limitation.
-For example, several participants asked SQL query strings, as these are typically constructed using user-generated data in practice.
-Although a simple expression entry box may suffice in simple scenarios, architectural support is needed for palettes that need to inspect the code context (e.g.
-to verify well-formedness) or if code highlighting, code completion and other advanced editing features are needed within the palette itself.
-
-\subsection{Keyboard Navigability (12)}
-
-Although our presentation of active code completion did not include any completely mouse-driven interfaces, several participants commented that the Color palette included interface elements taken from standard color dialog boxes that could only be manipulated using the mouse.
-These comments were generally severe in their condemnation of mouse-based interfaces in developer tools, consistent with our finding that a significant portion of our participants were familiar with editors like Vim that place a strong emphasis on keyboard shortcuts.
-
-\subsection{Minimal Impact on Editor Responsiveness}
-
-A common theme in our discussions with developers (as well as in comments left on our recruitment thread) was that integrated development environments like Eclipse were already too slow, and that an extension such as the one we were proposing would only be acceptable if it did not affect performance and responsiveness any further.
-
-\subsection{IDE and Language Portability}
-
-The mockups we showed users were based on the Java language and the Eclipse development environment.
-As we showed, a number of participants preferred other languages or editors.
-Many of these participants made comments asking that the features we describe not be restricted to one specific editor (e.g.
-NetBeans users) and some also called for support for other programming languages (e.g.
-Ruby and C\#).
-Indeed, the palettes we demonstrated could be used with only slight modifications in a variety of programming languages, given suitable architectural support for porting palettes between editing environments.
-
-\subsection{Varying User Needs}
-
-Our initial color palette was deemed overly complex by many participants (26), with many others (26) specifically saying that they would be satisfied with a simpler interface with colors listed next to their names.
-Other users wanted additional features, such as an ``eyedropper'' tool for selecting a color directly from an image on the screen.
-Similarly, the regular expression and SQL palettes were considered too simple by 12, and 15 participants respectively.
-User requests included syntax highlighting, a range of additional helpful tools for generating,  testing and sharing regular expressions, and the ability to browse databases more directly.
-Due to the wide range of user needs even within a single class, it may be that support for multiple palettes or a tabbed palette interface will be necessary in practice.
-To support incremental improvements to palettes based on user feedback, an architecture that makes deploying new and updated palettes relatively painless would also be valuable.
-
-\section{Use Cases}
-At the end of the survey, we solicited suggestions from the participants of other classes that could benefit from an associated palette.
-119 participants made one or more suggestion and we classified the suggestions into several categories.
-We also mention other suggestions made by researchers and developers via other channels in some cases, without including them in the provided counts.
-
-\subsection{Graphical Elements (27)}
-The most popular suggestions were graphical elements, influenced perhaps by our demonstration of the Color palette.
-Some participants suggested palettes for classes representing primitive graphical objects, such as brush and font selectors or polygon editors, while other participants were focused on user interface elements, such as buttons, check boxes and frame layouts.
-A few also suggested palettes for manipulating 3D primitives, such as transformation matrices, in a more direct and intuitive manner.
-A practitioner also suggested that because setting up a plot or graph is often significantly simpler using a direct manipulation interface, it would be a natural candidate for a palette as well.
-
-\subsection{Query Languages (17)}
-The second most popular category of suggestions consisted of various interfaces for query languages, also likely due to the examples we provided to participants.
-In addition to variants of the SQL and regular expression palettes, developers also wanted to work with other types of queries such as XPath or XQuery for XML.
-
-\subsection{Simplified or Domain-Specific Syntax (16)}
-Another interesting class of suggestions were cases where a more natural syntax than the syntax provided by Java is desirable.
-One suggestion was a palette that automatically escaped strings containing quotation marks or escape sequences.
-A related category of suggestions consisted of palettes that offered a more natural interface for generating strings containing code in other languages such as HTML (e.g.
-offering syntax highlighting, escaping,  tag matching and other features.) Domain-specific syntax for complex mathematical expressions and chemical formula were also mentioned in discussions with practitioners.
-
-An interesting suggestion that we investigated further involved Java's collection classes, such as \verb|ArrayList| and \verb|HashMap|.
-A participant suggested that these classes could be associated with a palette that offered a simplified literal syntax for initialization, pointing toward other languages that do offer such a literal syntax (e.g.
-JavaScript.) Without such syntax, these classes must be tediously initialized using a separate method call for each element.
-To determine whether this usage pattern is common, we conducted a corpus analysis using 10 randomly selected projects from the Qualitas Corpus \cite{QualitasCorpus:APSEC:2010} containing over 1M lines of code.
-We began by searching for places in these projects where Java collection classes were being instantiated, then looked to see whether this instantiation code was immediately followed by method calls that inserted items into the collection, indicating a case where a literal may have been used if available.
-Figure 4 summarizes the results of this analysis, providing evidence in support of the claim that a palette that simplifies this process could be useful for general-purpose programming.
-
-\subsection{Unclear Parameter Implications (11)}
-Another category of use cases contains classes where it can be difficult to predict what the run-time behavior of a particular parameter choice may be.
-Examples given included audio filters (e.g.
-pitch manipulation) and animation descriptors (e.g.
-speed or shape parameters).
-By giving immediate visual or auditory feedback using a preview panel, these parameters can be tweaked without requiring the execution of the full application.
-
+\subsubsection{Interaction with Code Context (13?)}
+
+Some people wanted to interact with the code context.
+For example, people expressed that it would be great if the SQL query palette was capable of dynamically creating SQL query string by combining multiple string variables, or by assigning variables for the parameters.
+
+\subsubsection{Performance (?)}
+
+Developers were concerned about the performance of these palettes.
+Some of the participants said that they would use these palettes only if it does not slow down the IDE.
+Also, this was the most popular comment on the reddit thread.
+
+\subsubsection{IDE Independence (?)}
+
+Several participants expressed a desire for IDE or even language independence for these palettes.
+
+\subsection{General Palette Design Considerations}
+
+General palette design considerations are about how to make more usable palettes.
+
+\subsubsection{Simplicity vs. Capability}
+
+Even though the participants were shown the same mockup screenshots, their reactions were split.
+Our color palette was considered too complex by many participants (26), and many others (26) would be satisfied with seeing colors next to a list of color names.
+The regular expressoin palette and SQL palette were considered too simple (12, and 15 participants respectively).
+They wanted syntax highlighting, match highlighting, and browsing capability of SQL databases.
+
+\subsubsection{Keyboard Navigability (12)}
+
+12 of the participants expressed strong antipathy toward the mouse. This was mostly directed at the color palette.
+
+\subsection{Suggestions from the Participants}
+At the end of the survey, we solicited suggestions from the participants about what types of classes could benefit from the active code completion palettes.
+119 participants suggested various types of classes and we classified the suggestions into several categories.
+
+\subsubsection{Alternative/Tricky/Literal Syntax (16)}
+These are the classes of which syntax is tricky, or complex.
+Even for a basic string class, people wanted to input multiline string or unicode string in more convenient way with automatic escaping.
+We address this issue in more detail in Section \#.
 %
@@
 
-\begin{figure}\label{collections}
-\begin{center}
-\begin{tabular}{l|c|c|r}
-Collection Class& Total & Literal & Percentage \\
-\hline
-\texttt{ArrayList} & 464	&	44	&	9.5\% \\
-\texttt{HashMap}	& 56 & 19 & 33.9\% \\
-\texttt{HashSet} & 122 & 62 & 50.8\% \\
-\texttt{Hashtable} & 86 & 10 & 11.6\% \\
-\texttt{Vector} & 729 & 31 & 4.2\% \\
-\hline
-\textbf{Total} & 1457 & 166 & 11.4\% \\
-\hline
-\end{tabular}
-\caption{Usage patterns for common Java collection classes in the \texttt{java.util} package in our code corpus.
-Uses that fit a pattern that can be captured by a literal make up a significant portion of all uses.
-Not all possible usage scenarios of this type were captured by our analysis, so these numbers are lower bounds.}
-\end{center}
-\end{figure}
+\subsubsection{Unclear Parameter Implications (11)}
+The classes in this category has some parameters in it, and it is not easy to predict the results of the parameter modifications.
+Instead of running the whole application to see the result, the palette can serve as a preview panel so that the developers may check the results directly.
+Also, the palette can be a control panel which facilitates modifying the parameters as in Juxtapose [?].
 
@@
 
+\subsubsection{Query Languages (17)}
+Similar to the regular expression and SQL palette we presented, the developers also wanted to check if the other types of query strings (e.g., XPath / XQuery) are correct or not by checking the results directly on the palette.
+
 %\begin{itemize}
@@
 
+\subsubsection{Graphical Elements (27)}
+The other popular category was graphical elements.
+We've already shown the color example, and the participants suggested many other graphical properties such as font, brush, shape and others.
+Also, people wanted to check or directly manipulate the GUI frames and layouts.
+
 %\begin{itemize}
@@
 
-\subsection{Integrating with Documentation and Examples (7)}
-Some participants suggested integrating tutorials or lists of relevant examples directly using a palette, so that these can be discovered more easily by new users and inserted directly into code, without requiring switching to a web browser and executing a search.
-
-\subsection{Complex Instantiation and Cleanup Procedures (5)}
-A related category contains classes that require complex instantiation and cleanup procedures.
+\subsection{Complex Instantiation Procedures (5)}
+Some of the participants pointed out that these palettes might be useful to create an object which requires complex instantiation code.
 For example, in order to read a text file in Java, the developer might want to use \texttt{BufferedReader} class.
-This class can be difficult to use because it requires try/catch block, and one must remember to close the file after reading it.
-By using a palette to choose a file or choose a variable which contains the file path, the developers could easily instantiate these objects and get an outline containing the full life-cycle of the file.
-Similarly, palettes may help to alleviate the factory pattern usability problem \cite{ellis_factory_2007}.
+It is error prone to use this class because it requires try/catch block, and also one must close the file after reading it.
+By using a palette to choose a file or choose a variable which contains the file path, the developers could easily instantiate these objects.
+In the same way, it can alleviate the factory pattern usability problem \cite{ellis_factory_2007}.
 As long as the developers remember which class to use, they will not need to remember how to instantiate that class.
-We explore this further in our user study in Section VI.
 
@@
 
-\subsection{Instantiation by Example (2)}
-In some cases, it is possible to describe an object by example.
-For instance, a class that represents a shortcut key combination may be most easily instantiated using a palette that simply reads a shortcut key from the developer.
-
-\subsection{Proof Assistants (1)}
-A proof assistant is a tool for constructing proof terms.
-According to the well-known Curry-Howard isomorphism between programming languages and formal logics, proof terms correspond to expressions and propositions correspond to types \cite{tapl}.
-Active code completion works directly with types to help developers construct expressions, so if applied to a language with a cleanly-developed connection to formal logic (e.g.
-Coq), palettes would be useful for constructing highly interactive proof assistant interfaces.
+\subsection{Describe by Example (2)}
+Sometimes, it is possible to describe an object by examples.
+For instance, if there is a class which represents a shortcut key combination, we can easily instantiate an object of that class by pressing the actual shortcut key on the interactive palette.
 
@@
 
-%\section{Tool}
-%Described earlier.
-%\section{Developer Survey}
-%
-%\subsection{Methods}
-%Our subject pool consisted primarily of visitors to a large collaborative filtering website dedicated to programming called {\tt /r/programming} on reddit.com\footnote{http://www.reddit.com/r/programming}.
-At the time of our study, there were approximately 340,000 registered members.
-A small number of participants were also recruited using a mass email to local computer science graduate students.
-The recruitment materials in both cases stated that we were seeking developers ''familiar with an object-oriented programming language like Java, C\# or Visual Basic and an integrated development environment like Eclipse or Visual Studio''.
-%
-%Participants were asked to take an online survey that would take approximately 20 minutes to complete.
-Of the 696 participants who started the survey, 475 participants finished it.
-Of these, 457 participants came from /r/programming and 18 from the graduate student mailing list.
-All quantitative metrics used in this study are based only on these participants.
-%
-%\subsection{Familiarity with Languages and Editors}
-%
-%We began by asking participants to rate their level of familiarity with several programming languages.
-Figure 1 summarizes these ratings.
-On average, participants rated themselves as very familiar with Java, C, C++ and JavaScript, familiar with C\#, Python and PHP and somewhat familiar with Visual Basic and Perl.
-%
-%We also asked participants to select the integrated development environments and code editors that they were familiar with.
-The Eclipse IDE was familiar to 87.1\% of participants.
-This was followed by Visual Studio with 66.0\% of participants, Vi/Vim with 53.7\%, Netbeans with 37.7\%, Emacs with 24.8\% and IntelliJ IDEA with 16.4\%.
-Participants could also enter ``other'' choices.
-A number of editors and IDEs were entered, including Xcode, Textmate, Notepad++ and others.
-%%\begin{figure}
-%%  \includegraphics[width=3.5in,height=2in]{experience.png}
-%%  \caption{Participant familiarity with various programming languages. Starting at bottom: None, Somewhat familiar, Familiar, Very Familiar, Expert.}
-%%\end{figure}
-%
-%\subsection{Mockups}
-%%\begin{figure*}
-%%  \includegraphics[scale=.6]{mockup-palette.png}
-%%  \caption{The regular expression palette, shown in Eclipse.}
-%%\end{figure*}
-%Next, we presented participants with a series of mockup palettes corresponding to a Color class, a regular expression Pattern class (Figure 2) and a SQL query ResultSet class.
-In addition to a screenshot of the palette itself, participants were shown mockups demonstrating how a user would invoke the palette, and a mockup showing the code that would be inserted once a selection had been made.
-%
-%Before presenting each palette mockup we gathered information about the user's familiarity with that topic and the strategies that they would likely use to instantiate that class.
-The Color class was presumed familiar to all participants.
-Figure 3 summarizes responses for the regular expression and SQL classes.
-The low number of participants unfamiliar with these topics again provides evidence that our participants were not generally novice developers.
-%
-%To determine the strategy typically used for instantiating the Color class, we asked the following question. The values in bold show the frequency of each response.
-%
-%\begin{quote}
-%A user interface designer asks you to change the background color of a window to ``navy blue'' when the user presses a button.
-You have determined that you need to instantiate the 'Color' class to do so.
-%\\\\
-%Which of the following strategies would you use first to create an instance of the 'Color' class representing navy blue?\\
-%
-%\begin{enumerate}[(a)]
-%\item I would use the IDE's content assist menu (see note below) to look for a predefined constant representing navy blue. (\textbf{58.4\%})
-%\item I would look up the documentation of the Color class to see if there is a predefined constant representing navy blue. (\textbf{19.0\%})
-%\item I would use an external tool (e.g. Photoshop or some other image editor) to determine the RGB values for navy blue. (\textbf{14.0\%})
-%\item I would guess at the RGB values for navy blue, then check whether my guess was correct by executing the application. (\textbf{5.3\%})
-%\item Other (please specify) (\textbf{3.3\%})\\
-%\end{enumerate}
-%\end{quote}
-%
-%Figure 4 summarizes the responses to a more general question asked about regular expressions and SQL queries.
-%
-%\begin{figure}
-%\begin{tabular}{lccc}
-% & \textbf{Regular Expressions} &  & \textbf{SQL}\\
-% \hline
-%Never used & 4.8\% &  & 0.0\%\\
-%Use infrequently & 46.7\% &  & 37.4\%\\
-%Use frequently & \textbf{48.4\%} &  & \textbf{62.6\%}\\
-%\hline
-%\end{tabular}
-%\caption{Participant's experience with regular expressions and SQL.}
-%\end{figure}
-%
-%\begin{figure}
-%\begin{tabular}{lccc}
-% & \textbf{Regular Expressions} &  & \textbf{SQL}\\
-% \hline
-%Separate test script & 29.7\% &  & 15.5\%\\
-%Guess and check & 13.4\% &  & 16.0\%\\
-%External tool & \textbf{38.5\%} &  & \textbf{58.9\%}\\
-%Search for examples & 12.1\% & & 4.8\% \\
-%Other & 6.2\% & & 4.8\% \\
-%\hline
-%\end{tabular}
-%\caption{Typical strategies for regular expressions and SQL queries.}
-%\end{figure}
-%
-%
-%%For each of these classes,
-%
-%
-%%\subsubsection{Color}
-%%We began with a class representing colors. We asked participants to consider the following scenario:
-%
-%\begin{figure}
-%\begin{tabular}{crccccc}\\\\
-%\textsc{class}
-%& 
-%& \begin{rotate}{20}Nearly every time\end{rotate}
-%& \begin{rotate}{20}Most of the time\end{rotate}
-%& \begin{rotate}{20}Some of the time\end{rotate}
-%& \begin{rotate}{20}Rarely\end{rotate}
-%& \begin{turn}{20}Never\end{turn}\\
-%\hline
-%Color &\vline& 9.6\% & 22.1\% & \textbf{32.4\%} & 28.2\% & 7.7\%\\
-%RegExp &\vline& \textbf{36.6\%} & 29.5\% & 21.8\% & 7.3\% & 4.8\%\\
-%SQL & \vline &18.2\% & 19.3\% & \textbf{30.9\%} & 20.4\% & 11.4\%\\
-%\hline
-%\end{tabular}
-%\caption{For each of the classes considered, participants were shown mockups of a code completion palette designed for instantiating that class, as well as the source code generated after parameters were entered.
-This table shows the distribution of responses to the question: ``Consider situations where you need to instantiate the [specified] class.
-What portion of the time, in these situations, do you think you would use this feature?''}
-%\end{figure}
-%%
-%%\begin{figure}
-%%  \includegraphics[scale=.6]{regex-insert.png}
-%%  \caption{Test}
-%%\end{figure}
-%%
-%%
-%
-%%Then we asked:
-%%\begin{quote}
-%%\textbf{Q2} Consider situations where you need to instantiate the 'Color' class. What portion of the time, in these situations, do you think you would use this feature?
-%%\begin{enumerate}[(a)]
-%%\item Nearly every time (\textbf{9.6\%})
-%%\item Most of the time (\textbf{22.1\%})
-%%\item Some of the time (\textbf{32.4\%})
-%%\item Rarely (\textbf{28.2\%})
-%%\item Never (\textbf{7.7\%})\\
-%%\end{enumerate}
-%%\end{quote}
-%%
-%
-%After presenting each palette, we asked users to rate the usefulness of each palette.
-Responses are summarized in Figure 5.
-We then allowed participants to provide qualifications and open-ended feedback by asking the following question: ``If you would like to qualify or elaborate on your answer to the previous question you may do so below.
-You may also suggest improvements for this palette.'' This information is analyzed in the remainder of the paper.
-%
-%%\begin{quote}
-%%\textbf{Q4} How would you describe your level of familiarity with (regular expressions, SQL)?\\
-%%\\
-%%\begin{enumerate}[(a)]
-%%\item I do not know what [it is].	 (\textbf{0.7\%}, \textbf{0.0\%})
-%%\item I know what regular expressions are but I have not used them in practice. (\textbf{4.4\%, 0.0\%})
-%%\item I use regular expressions infrequently.	 (\textbf{46.0\%, 35.9\%})
-%%\item I use regular expressions frequently. (\textbf{49.0\%, 64.1\%})
-%%\end{enumerate}
-%%\end{quote}
-%
-%
-%
-%%\begin{quote}
-%%\textbf{Q5} Consider situations where you need to write a regular expression to extract information from a string or collection of strings. What strategy would you most likely use first?
-%
-%%\begin{enumerate}[(a)]
-%%\item I would write a separate test script to help come up with an appropriate regular expression. (\textbf{29.7\%})
-%%\item I would guess and check regular expressions by running my application. (\textbf{13.4\%})
-%%\item I would launch an external application or web tool to help me test regular expressions interactively. (\textbf{38.5\%})
-%%\item I would search the web for a sample regular expression suitable for my task. (\textbf{12.1\%})
-%%\item Other (please specify) (\textbf{6.2\%})
-%%\end{enumerate}
-%%\end{quote}
-%
-%
-%%\begin{quote}
-%%\textbf{Q6} Consider situations where you need to create a regular expression. What portion of the time, in these situations, do you think you would use this feature?
-%%\begin{enumerate}[(a)]
-%%\item Nearly every time (\textbf{36.6\%})
-%%\item Most of the time (\textbf{29.5\%})
-%%\item Some of the time (\textbf{21.8\%})
-%%\item Rarely (\textbf{7.3\%})
-%%\item Never (\textbf{4.8\%})\\
-%%\end{enumerate}
-%%\end{quote}
-%
-%%\begin{quote}
-%%\textbf{Q7} How would you describe your level of familiarity with SQL?
-%
-%%\begin{enumerate}[(a)]
-%%\item I do not know what regular expressions are.	 (\textbf{0.0\%})
-%%\item I know what SQL is I have not used it in practice. (\textbf{0.0\%})
-%%\item I use SQL infrequently.	 (\textbf{35.9\%})
-%%\item I use SQL frequently. (\textbf{64.1\%})\\
-%%\end{enumerate}
-%%\end{quote}
-%
-%%\begin{quote}
-%%\textbf{Q8} Consider situations where you need to write a regular expression to extract information from a string or collection of strings. What strategy would you most likely use first?
-%
-%%\begin{enumerate}[(a)]
-%%\item I would write a separate test script to help come up with an appropriate SQL query.	
-% %(\textbf{15.5\%})
-%%\item I would guess and check by running my application. (\textbf{16.0\%})
-%%\item I would launch an external application or web tool to test my query against my database interactively.	(\textbf{58.9\%})
-%%\item I would search the web for an example SQL query suitable for my task.	
-% %(\textbf{4.8\%})
-%%\item Other (please specify) (\textbf{4.8\%})
-%%\end{enumerate}
-%%\end{quote}
-%
-%%Showed the SQL palette and asked:
-%
-%%\begin{quote}
-%%\textbf{Q9} Consider situations where you need to create a SQL query. How often, in these situations, do you think you would use this feature?
-%%\end{quote}
-%
-%\section{Design Implications}
-%%\begin{figure}[b]
-%%  \includegraphics[scale=.5]{color-palette.png}
-%%  \caption{test}
-%%\end{figure}
-%
-%We next analyzed the open-ended responses to extract some design principles and constraints.
-We got 193 responses for the color palette, 129 for the regular expression palette, and 142 for the SQL palette.
-Also, we got total 119 suggestions about what other types of classes could benefit from similar types of palettes.
-The design can be divided into two main categories: system design constraints, and general palette design considerations.
-System design constraints are the criteria that should be taken into account in the design of the overall system architecture, and the general palette design considerations are about how to make more usable palettes.
-%
-%\subsection{System Design Constraints}
-%
-%\subsubsection{Handling Separation of Concerns}
-%
-%Professional developers were very wary of including constant data into the source code directly.
-Many responses expressed general sentiment or specific preference for project-wide color theme or external resource file.
-Also, many people stated that these tools might not be useful since they usually read an external resource file, or sometimes they explicitly suggested to add capability to handle the external resource files directly from the palette.
-%
-%Examples of feedbacks in this category include:
-%
-%\begin{itemize}
-%	\item The use of magic numbers in the source code is a bad design (54)
-%	\item Color picking is the `designer's job' (33)
-%	\item Would only use to define named constants (17)
-%	\item Would rather use color scheme / theme (3)
-%	\item Would use Object-Relational Mapping (ORM) / Language INtegrated Query (LINQ) instead of raw SQL query string (27 of 142)
-%	\item These palettes might be useful for prototyping, but not for real production (14)
-%	\item Instead of including the test cases into comments, they should be stored as unit tests (35 of 143)
-%\end{itemize}
-%
-%Interestingly, most people seemed to consider regular expressions as part of the program logic.
-There were few people who complained about including the regular expression pattern string in the source code.
-% 
-%\subsubsection{Reversibility}
-%
-%19 participants expressed that it should be able to bring the palettes back when they want to modify the parameters after they used the palettes.
-% 
-%\subsubsection{Palette Settings and State}
-%
-%Many participants wanted to be able to configure the palettes or have the palettes maintain state even after the palettes are hidden.
-The followings are the comments that are related to this category.
-%
-%\begin{itemize}
-%	\item Want recently used regular expressions, and control over the comments (12)
-%	\item Persistent database connection information is needed / want to manage connection pools (9)
-%	\item Want recent colors, capability of setting favorite colors (20)
-%\end{itemize}
-%
-%\subsubsection{Interaction with Code Context}
-%
-%Some people wanted to interact with the code context.
-For example, 13 people expressed that it would be great if the SQL query palette was capable of dynamically creating SQL query string by combining multiple string variables, or by assigning variables for the parameters.
-%
-%\subsubsection{Performance}
-%
-%Developers were concerned about the performance of these palettes.
-Some of the participants said that they would use these palettes only if it does not slow down the IDE.
-Also, this was the most popular comment on the reddit thread.
-%
-%\subsubsection{IDE Independence}
-%
-%Several participants expressed a desire for IDE or even language independence for these palettes.
-%
-%%\subsubsection{Composing palette logic}
-%
-%\subsection{General Palette Design Considerations}
-%\subsubsection{Simplicity vs. Capability}
-%
-%Even though the participants were shown the same mockup screenshots, their reactions were split.
-Our color palette was considered too complex by many participants (26), and many others (26) would be satisfied with seeing colors next to a list of color names.
-The regular expressoin palette and SQL palette were considered too simple (12, and 15 participants respectively).
-They wanted syntax highlighting, match highlighting, and browsing capability of SQL databases.
-%
-%\subsubsection{Keyboard Navigability}
-%
-%12 of the participants expressed strong antipathy toward the mouse. This was mostly directed at the color palette.
-%
-%\subsubsection{Complexity}
-%
-%%\subsubsection{Side Effects}
-%
-%\section{Suggestions from the Participants}
-%We classified the suggestions into several categories. The number in the parentheses indicate the number of participants who suggested the classes which fall into the category.
-%
-%\subsection{Alternative/Tricky/Literal Syntax (16)}
-%These are the classes of which syntax is tricky, or complex. Even for a basic string class, people wanted to input multiline string or unicode string in more convenient way with automatic escaping.
-%
-%\begin{itemize}
-%	\item Example classes
-%	
-%	\begin{itemize}
-%		\item String
-%		\item Collections (e.g., dictionary)
-%		\item Vector, Matrix
-%		\item Embedded languages
-%		\item URLs, Paths
-%	\end{itemize}
-%	
-%	\item Expected features
-%	
-%	\begin{itemize}
-%		\item Proper escaping
-%		\item Syntax highlighting
-%		\item Auto-completion
-%	\end{itemize}
-%\end{itemize}
-%
-%\subsection{Unclear Parameter Implications (11)}
-%The classes in this category has some parameters in it, and it is not easy to predict the results of the parameter modifications.
-Instead of running the whole application to see the result, the palette can serve as a preview panel so that the developers may check the results directly.
-Also, the palette can be a control panel which facilitates modifying the parameters.
-%\begin{itemize}
-%	\item Example classes
-%	
-%	\begin{itemize}
-%		\item Audio tweaks (modifying pitch, volume, etc.)
-%		\item 3D transformation matrices
-%		\item number / string / date formatting
-%		\item message box / input box (e.g., \texttt{JOptionPane})
-%	\end{itemize}
-%\end{itemize}
-%
-%\subsection{Query Languages (17)}
-%Similar to the unclear parameter implications category, the developers wanted to check if the query string is correct or not by checking the results directly on the palette.
-%
-%\begin{itemize}
-%	\item Example classes
-%	
-%	\begin{itemize}
-%		\item Regular expression
-%		\item SQL query
-%		\item XPath / XQuery
-%	\end{itemize}
-%\end{itemize}
-%
-%\subsection{Graphical Elements (27)}
-%The other popular category was graphical elements.
-We've already shown the color example, and the participants suggested many other graphical properties such as font, brush, shape and others.
-Also, people wanted to check or directly manipulate the GUI frames and layouts.
-%
-%\begin{itemize}
-%	\item Example classes
-%	
-%	\begin{itemize}
-%		\item Color, Shape, Brush, Font, etc.
-%		\item JFrame, Layouts, any swing components
-%	\end{itemize}
-%\end{itemize}
-%
-%\subsection{Complex Instantiation Procedures (5)}
-%Some of the participants pointed out that these palettes might be useful to create an object which requires complex instantiation code.
-For example, in order to read a text file in Java, the developer might want to use \texttt{BufferedReader} class.
-The instantiation code might look as following:
-%\begin{verbatim}
-%BufferedReader reader = null;
-%try {
-%    reader = new BufferedReader(
-%        new FileReader(filePath));
-%} catch (IOException e) {
-%    e.printStackTrace();
-%}
-%// use reader here
-%reader.close();
-%\end{verbatim}
-%By using a palette to choose a file or choose a variable which contains the file path, the developers could easily instantiate these objects.
-In the same way, it can alleviate the factory pattern usability problem.
-As long as the developers remember which class to use, they will not need to remember how to instantiate that class.
-%
-%\begin{itemize}
-%	\item Example classes
-%	
-%	\begin{itemize}
-%		\item BufferedReader
-%		\item Classes using factory patterns
-%		\item Database connection string builder
-%	\end{itemize}
-%\end{itemize}
-%
-%\subsection{Describe by Example (2)}
-%Sometimes, it is possible to describe an object by examples.
-For instance, if there is a class which represents a shortcut key combination, we can easily instantiate an object of that class by pressing the actual shortcut key on the interactive palette.
-%
-%\begin{itemize}
-%	\item Example classes
-%	
-%	\begin{itemize}
-%		\item Keyboard Keys
-%		\item Regular expression
-%	\end{itemize}
-%\end{itemize}
-%
-%\subsection{Integrating with Documentation, Tutorial (7)}
-%Some participants suggested integrating the documentation or the tutorials into the interactive palettes.
+\subsection{Integrating with Documentation, Tutorial (7)}
+Some participants suggested integrating the documentation or the tutorials into the interactive palettes.
 If the developers are not familiar of the specific API class, the palette could tell them how to use it.
-%
-%%\begin{itemize}
-%%\item 54 expressed a concern about encouraging inlining of constants, preferring external resource files.
-17 stated that they would use this feature only to define constants.
-33 stated that it should be the designer's job.
-%%\item 11 expressed a concern about keyboard navigability
-%%\item 27 expressed a concern about the complexity of the palette
-%%\item 10 expressed an interest in a color picker. 7 wanted a multitabbed window with different options for picking colors.
-%%\item 26 expressed a desire for color names, 10 of which wanted to ensure color constant names were used if available
-%%\item 20 wanted persistent state (MRU, favorites)
-%%\end{itemize}
-%
-%% An example of a floating figure using the graphicx package.
-%% Note that \label must occur AFTER (or within) \caption.
-%% For figures, \caption should occur after the \includegraphics.
-%% Note that IEEEtran v1.7 and later has special internal code that
-%% is designed to preserve the operation of \label within \caption
-%% even when the captionsoff option is in effect. However, because
-%% of issues like this, it may be the safest practice to put all your
-%% \label just after \caption rather than within \caption{}.
-%%
-%% Reminder: the "draftcls" or "draftclsnofoot", not "draft", class
-%% option should be used if it is desired that the figures are to be
-%% displayed while in draft mode.
-%%
-%%\begin{figure}[!t]
-%%\centering
-%%\includegraphics[width=2.5in]{myfigure}
-%% where an .eps filename suffix will be assumed under latex, 
-%% and a .pdf suffix will be assumed for pdflatex; or what has been declared
-%% via \DeclareGraphicsExtensions.
-%%\caption{Simulation Results}
-%%\label{fig_sim}
-%%\end{figure}
-%
-%% Note that IEEE typically puts floats only at the top, even when this
-%% results in a large percentage of a column being occupied by floats.
-%
-%
-%% An example of a double column floating figure using two subfigures.
-%% (The subfig.sty package must be loaded for this to work.)
-%% The subfigure \label commands are set within each subfloat command, the
-%% \label for the overall figure must come after \caption.
-%% \hfil must be used as a separator to get equal spacing.
-%% The subfigure.sty package works much the same way, except \subfigure is
-%% used instead of \subfloat.
-%%
-%%\begin{figure*}[!t]
-%%\centerline{\subfloat[Case I]\includegraphics[width=2.5in]{subfigcase1}%
-%%\label{fig_first_case}}
-%%\hfil
-%%\subfloat[Case II]{\includegraphics[width=2.5in]{subfigcase2}%
-%%\label{fig_second_case}}}
-%%\caption{Simulation results}
-%%\label{fig_sim}
-%%\end{figure*}
-%%
-%% Note that often IEEE papers with subfigures do not employ subfigure
-%% captions (using the optional argument to \subfloat), but instead will
-%% reference/describe all of them (a), (b), etc., within the main caption.
-%
-%
-%% An example of a floating table. Note that, for IEEE style tables, the 
-%% \caption command should come BEFORE the table. Table text will default to
-%% \footnotesize as IEEE normally uses this smaller font for tables.
-%% The \label must come after \caption as always.
-%%
-%%\begin{table}[!t]
-%%% increase table row spacing, adjust to taste
-%%\renewcommand{\arraystretch}{1.3}
-%% if using array.sty, it might be a good idea to tweak the value of
-%% \extrarowheight as needed to properly center the text within the cells
-%%\caption{An Example of a Table}
-%%\label{table_example}
-%%\centering
-%%% Some packages, such as MDW tools, offer better commands for making tables
-%%% than the plain LaTeX2e tabular which is used here.
-%%\begin{tabular}{|c||c|}
-%%\hline
-%%One & Two\\
-%%\hline
-%%Three & Four\\
-%%\hline
-%%\end{tabular}
-%%\end{table}
-%
-%
-%% Note that IEEE does not put floats in the very first column - or typically
-%% anywhere on the first page for that matter. Also, in-text middle ("here")
-%% positioning is not used. Most IEEE journals/conferences use top floats
-%% exclusively. Note that, LaTeX2e, unlike IEEE journals/conferences, places
-%% footnotes above bottom floats. This can be corrected via the \fnbelowfloat
-%% command of the stfloats package.
-%
-%
-%
-%\section{Conclusion}
-%Overall, professional developers appear willing to use these for a number of use cases.
-However, the usability criteria described above impose stiff requirements on the architecture.
-Of particular concern, we found that the developers preferred help with program logic, but not the constant data.
-This indicates that another area of application may be in specialized editors for resource files.
-%
-%
-\section{Tool Design}
-Motivated by these use cases and with the design criteria developed as a result of our developer survey in mind, we sought to create an active code completion system.
-We named our project Graphite, an acronym for {\bf Gra}phical {\bf P}alettes to {\bf H}elp {\bf I}nstantiate {\bf T}ypes in the {\bf E}ditor.
-We describe the design decisions and trade-offs that went into this system in this section.
-A video demonstrating Graphite, including the process of writing a full palette, is available for further details \footnote{http://www.youtube.com/watch?v=FfCoXQEHGAI}.
-The full source code of Graphite is also available at the URL given on the first page.
-
-We chose to use Eclipse for the Java programming language because this combination was the most widely-used  amongst participants in our survey.
-Eclipse plug-in development is well-supported and there are many examples available online that we can refer to, and the Eclipse project is available under an open source license.
-
-\subsection{HTML+Javascript based palettes}
-Eclipse is itself written in Java and uses the SWT and JFace graphical user interface toolkits.
-These frameworks are not widely used outside of the Eclipse ecosystem.
-As we have noted, a number of participants in our field study indicated that they hoped that our tool would be available for other IDEs and other programming languages.
-Indeed, many concepts (including regular expressions, colors and database queries) are similar across languages.
-To better support IDE-independence and language-independence, we chose instead to build palettes using HTML and Javascript.
-Javascript was among the most well-known languages in our study, just behind Java, C and C++, and is highly flexible.
-A number of user interface libraries are available, such as jQuery\footnote{http://jquery.com/} (which we used to implement our palettes).
-All major browsers now feature sophisticated debuggers and run-time inspection facilities that ease debugging and testing.
-Deploying palettes using standard URLs is also simpler than attempting to integrate their source code directly into Java libraries and packages and eases the process of incremental development and deployment.
-All major windowing toolkits feature a web browser control that can allow for the display of these palettes within the editor environment.
-In Eclipse, we used the standard SWT Browser control.
-
-\subsection{Palette API}
-In order to allow palettes to communicate with the surrounding editor environment, a simple Javascript interface was injected.
-These functions were implemented by the editor plug-in, but appear as ``native'' browser functions to Javascript code.
-The API consists of the following methods, accessed through a global object named \verb|graphite|:
-\begin{itemize}
-	\item \verb|insert(str)|
-
-	Inserts the specified string at the cursor and closes the palette. The appropriate amount of indentation is automatically inserted after any newlines in this string.
-
-	\item \verb|cancel()|
-
-	Closes the palette without inserting a string.
-
-	\item \verb|getSelectedText()|
-
-	Returns the text that is currently selected by the user in the editor, or an empty string if no text has been selected. This method is used to implement reinvocation.
-
-	\item \verb|getIDE()|
-
-	Returns a string that specifies the IDE that is being used, ``Eclipse'' in our implementation.
-
-	\item \verb|getLanguage()|
-
-	Returns a string that specifies the programming language that is being used, ``java'' in our implementation.
-\end{itemize}
-
-By limiting the complexity of this API, we reasoned that developers would be able to create specialized palettes more easily.
-An additional benefit is that the developers of other editing environments are also able to create plug-ins that support Graphite palettes with minimal effort.
-
-To access this API, the palette must include a small script named \verb|graphite.js| into their page.
-This file interacts with the host editor to create the graphite object and provides an alternative implementation of these methods when the palette is being tested in a web browser.
-
-\subsection{Palette Discovery}
-\begin{figure*}\begin{center}
-\includegraphics[scale=.6]{regex.png}\end{center}
-\caption{The regular expression palette developed using Graphite and used by subjects in the treatment group of the user study described in Sec. VI.}
-\end{figure*}
-
-Graphite currently provides two methods for associating a palette with a particular type so that the Graphite plug-in is able to discover it when the code completion menu is invoked by a developer.
-
-\subsubsection{Annotation-based}
-For user-defined classes that the palette author has the authority to modify, the \texttt{@GraphitePalette} annotation associates a palette with the class.
-The annotation must specify the URL of the palette, and can contain some other optional information (e.g.
-the description that is shown in the code completion menu).
-This  allows library providers to provide palettes that are specialized to their libraries and distribute them alongside their code.
-In this way, the users of the libraries are not required to discover that an external tool exists.
-If the user does not have the Graphite plug-in for Eclipse installed, the annotation has no effect.
-
-\subsubsection{Explicit}
-In cases where the palette developer cannot modify a class directly (such as palettes for classes in the Java standard library), a second method exists to explicitly associate a palette with the fully qualified name of a class.
-Currently, this feature is exposed as a preference pane in the Eclipse IDE.
-To facilitate automated installs in the future, a preferences file may also be added to the system.
-
-\subsection{Design Trade-Offs}
-The design that we have described is light-weight, highly flexible and highly responsive.
-It lays foundations for IDE and language portability.
-However, several trade-offs have also been made to achieve these benefits.
-
-\begin{itemize}
-\item Because palettes are implemented in Javascript, any differences between Java and Javascript become evident.
-For example, Javascript color names are, in two cases, slightly different from the corresponding Java color name.
-The Javascript regular expression engine is also slightly different.
-Although a Java applet could be embedded to provide necessary interoperability in cases where these differences are critical, this is more difficult than it would be if the palettes were implemented using the same language as the code.
-\item The palette user interface must stay within its bounding box -- user interface elements like pop-up menus (unless they are provided by the browser itself) are thus more difficult to implement and may require additional API support in the future.
-\item Several use cases could benefit from greater access to the surrounding code, or even the surrounding project.
-Implementing this modularly is particularly difficult, and the Eclipse API does not easily allow for  serialization of the sum of its contextual knowledge for consumption by a palette.
-\item Implementing a reinvocation mechanism that is robust to changes in the code and does not require the user to select the relevant information is difficult because this would require that the palette  parse all the code to find relevant information.
-This also would require richer insertion APIs that allow for  ``dead'' code to be removed.
-This puts a greater burden on palette developers and errors could affect user's code in subtle ways.
-The small inconvenience of having to select code before reinvoking a palette may be balanced by the knowledge that no other portions of the code will be modified inadvertently by that palette.
-\end{itemize}
-
-\subsection{Palettes}
-We implemented two full-featured palettes to demonstrate and test the functionality of this framework: a regular expression palette (Figure 5) and a color selection palette (Figure 1).
-
-\subsubsection{Regular Expressions}
-Writing correct regular expression patterns is difficult.
-The regular expression palette, associated with the \verb|Pattern| class in the Java standard library, allows users to enter and test regular expression patterns interactively before inserting them into their code.
-The focal point of this palette is the pattern input area.
-As the user enters a pattern into this input area, syntax errors are indicated with a red background; the background remains white when the pattern is valid.
-
-In addition to the pattern, a regular expression consists of flags that can change its matching behavior in various ways.
-Our palette allows users to toggle the case-sensitivity flag of the regular expression using a small checkbox labeled \verb|Ignore Case|, placed next to the input area.
-A keyboard shortcut is also available, indicated using the underlined letter (Ctrl+I).
-
-To allow users to test the behavior of the regular expression that they have entered, two columns are available below the expression input area.
-The left column contains an input area labeled ``Should match...'' and the right column contains an input area labeled ``Should NOT match...''.
-Users use these to enter lists of test  strings into each column.
-The background colors behind these strings change to indicate whether the regular expression that has been entered matches or does not match that string.
-Green indicates that the pattern  {\it matched} the string and red indicates that it does {\it not} match, regardless of the column that the test is in (this scheme was chosen based on feedback from an initial pilot study of this palette.) A key describing this color scheme is displayed after the first test has been entered (not shown).
-
-Users navigate between the three input areas using standard \verb|Tab| cycling behavior.
-The label in each text area remains visible until some input has been entered, rather than disappearing immediately on focus.
-When a user is satisfied with the regular expression that they have entered, she can press the \verb|Enter| key to insert the appropriate Java source code.
-Because Java requires that the regular expression pattern be placed inside a string literal, additional escape sequences are needed in front of backslashes.
-This can be tedious and error-prone if done manually.
-Because the palette is responsible for generating this literal, however, this is no longer a problem.
-In addition to the source code itself, the tests are retained in a comment beginning on the next line.
-The user can leave these as documentation or use them to create more formal unit tests within their preferred testing framework.
-
-If the user wishes to modify the regular expression or change the test set, she can highlight the code that was inserted and then invoke the palette once again.
-The palette parses the selected text to extract the regular expression and tests.
-Although this is slightly inconvenient in that it requires that the user select the relevant portion of the code, it is the most robust method for implementing this feature -- it can easily handle changes to the regular expression or tests and it does not require additional logic in the palette to determine which portions of the code are relevant, and should subsequently be replaced if a new regular expression is entered.
-
-\subsubsection{Color Selection}
-We omit a description of the full behavior of the Color selection palette due to space limitations. 
-%\section{Future work}
-%
-%\subsection{Graphite plug-in for different IDEs}
-%As all the palettes are independent to Eclipse, We speculate that implementing Graphite plug-in for other IDEs will not be too difficult.
-Implementing a web-based aiding tool and sharing that tool among different IDEs with minimal effort will be a good idea.
-%
-%\subsection{Improving the Graphite API for Javascript}
-%The current version of Graphite API for Javascript is not powerful enough.
-For example, the palettes cannot add comments before current line, nor can it delete or replace code fragment which was not highlighted by the user.
-By providing more effective and powerful API, the palette writers can write the palettes more quickly and make more useful palettes.
-%
-%\subsection{More palette examples}
-%Due to the limited time, we could only provide two example palettes so far. More palette examples for various types might inspire the library providers and palette writers to design better palettes.
-%
-%\subsection{Keyword invocation model / online palette repository}
-%Although Keyword invocation has many expected benefits, it is currently not supported.
-This model will make Graphite language-independent, and improve discoverability of the palettes by allowing users to search the palette with keywords.
-Palette search will work better if a shared online palette repository is built and maintained.
-%
-%
-
-% An example of a floating figure using the graphicx package.
-% Note that \label must occur AFTER (or within) \caption.
-% For figures, \caption should occur after the \includegraphics.
-% Note that IEEEtran v1.7 and later has special internal code that
-% is designed to preserve the operation of \label within \caption
-% even when the captionsoff option is in effect. However, because
-% of issues like this, it may be the safest practice to put all your
-% \label just after \caption rather than within \caption{}.
-%
-% Reminder: the "draftcls" or "draftclsnofoot", not "draft", class
-% option should be used if it is desired that the figures are to be
-% displayed while in draft mode.
-%
-%\begin{figure}[!t]
-%\centering
-%\includegraphics[width=2.5in]{myfigure}
-% where an .eps filename suffix will be assumed under latex, 
-% and a .pdf suffix will be assumed for pdflatex; or what has been declared
-% via \DeclareGraphicsExtensions.
-%\caption{Simulation Results}
-%\label{fig_sim}
-%\end{figure}
-
-% Note that IEEE typically puts floats only at the top, even when this
-% results in a large percentage of a column being occupied by floats.
-
-% An example of a double column floating figure using two subfigures.
-% (The subfig.sty package must be loaded for this to work.)
-% The subfigure \label commands are set within each subfloat command, the
-% \label for the overall figure must come after \caption.
-% \hfil must be used as a separator to get equal spacing.
-% The subfigure.sty package works much the same way, except \subfigure is
-% used instead of \subfloat.
-%
-%\begin{figure*}[!t]
-%\centerline{\subfloat[Case I]\includegraphics[width=2.5in]{subfigcase1}%
-%\label{fig_first_case}}
-%\hfil
-%\subfloat[Case II]{\includegraphics[width=2.5in]{subfigcase2}%
-%\label{fig_second_case}}}
-%\caption{Simulation results}
-%\label{fig_sim}
-%\end{figure*}
-%
-% Note that often IEEE papers with subfigures do not employ subfigure
-% captions (using the optional argument to \subfloat), but instead will
-% reference/describe all of them (a), (b), etc., within the main caption.
-
-% An example of a floating table. Note that, for IEEE style tables, the 
-% \caption command should come BEFORE the table. Table text will default to
-% \footnotesize as IEEE normally uses this smaller font for tables.
-% The \label must come after \caption as always.
-%
-%\begin{table}[!t]
-%% increase table row spacing, adjust to taste
-%\renewcommand{\arraystretch}{1.3}
-% if using array.sty, it might be a good idea to tweak the value of
-% \extrarowheight as needed to properly center the text within the cells
-%\caption{An Example of a Table}
-%\label{table_example}
-%\centering
-%% Some packages, such as MDW tools, offer better commands for making tables
-%% than the plain LaTeX2e tabular which is used here.
-%\begin{tabular}{|c||c|}
-%\hline
-%One & Two\\
-%\hline
-%Three & Four\\
-%\hline
-%\end{tabular}
-%\end{table}
-
-% Note that IEEE does not put floats in the very first column - or typically
-% anywhere on the first page for that matter. Also, in-text middle ("here")
-% positioning is not used. Most IEEE journals/conferences use top floats
-% exclusively. Note that, LaTeX2e, unlike IEEE journals/conferences, places
-% footnotes above bottom floats. This can be corrected via the \fnbelowfloat
-% command of the stfloats package.
-
-\section{User Study Design}
-
-We conducted a comparative study to evaluate the usefulness and usability of a Graphite palette in a constrained setting -- writing regular expressions.
-We used a between-subjects design by randomly splitting the participants into two groups: a control group, where subjects were not shown nor able to use the palette, and a treatment group, where subjects were shown a different palette and allowed to discover and use the regular expression palette if they wished.
-
-\subsection{Process}
-\subsubsection{Preliminary Survey}
-The participants were first asked to fill out the pre-survey form about their prior experiences with various programming languages, integrated development environments (IDEs), and regular expressions, based on the previous survey that we had conducted.
-Of note, the subjects in this study were considerably less skilled with Java and regular expressions than in the previous survey, likely due to a sampling effect -- our subjects were all PhD students at Carnegie Mellon.
-
-\subsubsection{Training}
-Only the participants in the treatment group were shown how to invoke Graphite palettes in the context of an  Eclipse code editor with a palette for the Color class.
-We chose to demonstrate the tool using a color palette instead of the regular expression palette itself because we wanted to simulate the condition where a user had discovered the palette naturally.
-The demonstration of the Color palette was brief, taking about two minutes.
-We then described the nature of the task to the subject and allowed them to begin, giving them 45 minutes to complete all tasks, in any order.
-
-\subsection{Within vs. between-subjects design}
-
-We decided to use a \textit{between-subjects design} because a within-subjects design would have required us to produce pairs of tasks with equal difficulties.
-This turned out to be very difficult.
-Second, we could not easily ignore a learning effect during the experiment if we had used a within-subjects design.
-We observed that this effect was quite strong, as most subjects hadn't used regular expressions recently.
-
-\subsection{Participants}
-
-We recruited 7 subjects from Computer Science Department and Institute for Software Research at Carnegie Mellon University.
-All the 7 subjects were PhD students.
-The subjects were randomly assigned into two groups: 4 subjects were assigned to the control group, and the other 3 subjects to the treatment group.
-There were six male participants and one female participant.
-Participants were compensated in the amount of 15 dollars for their participation.
-
-\subsection{Tasks}
-
-There were a total of 9 tasks to be completed in 45 minutes.
-The first 6 tasks involved writing regular expressions to validate various data formats (e.g.
-temperatures), and the remaining 3 tasks involved writing regular expressions to retrieve data from a document.
-The participants were allowed to move back and forth while doing the experiment, and they were allowed to use any external resources, including the internet, local console, and so forth (we did not observe any usage of programs other than the web browser, however).
-The only restriction we placed on their activity was that they were NOT allowed to directly search for the answer to a task online online.
-We omit descriptions of each individual question due to space limitations.
-
-%
-%\begin{itemize}
-%	\item Data format validation tasks (total 6 tasks)
-%	
-%	\begin{enumerate}
-%		\item US phone numbers
-%		
-%		For simplicity, it was assumed that any digit can be used in any place in the phone number, even though there is a notion of valid range for the area code.
-%		
-%		\item Times
-%		
-%		Hours and minutes only.
-%		
-%		\item Temperatures
-%		
-%		Only Fahrenheit and Celcius were used, and subjects were instructed to assume the degree symbol had not been used.
-%		
-%		\item Dates
-%		
-%		We did not place any restrictions on the format for dates.
-This task is quite difficult to solve generally.
-There are many valid formats for expressing dates, and the valid range of the day part depends on the specific year and month.
-Many participants were stuck on this task and chose to skip it and came back later.
-%		
-%		\item IPv4 addresses
-%		
-%		\item URLs starting with \texttt{http://} or \texttt{https://}
-%		
-%	\end{enumerate}
-%	
-%	\item Data searching tasks (total 3 tasks)
-%	
-%	\begin{enumerate}
-%	\setcounter{enumi}{6}
-%		\item Finding a fields declaration (member variable) in a Java class
-%		
-%		It was assumed that all the field declarations start with one of the access modifier keywords: \verb|public|, \verb|protected|, or \verb|private|.
-%		
-%		\item Finding a string literal in a Java file
-%		
-%		\item Finding an \texttt{img} tag in an HTML file
-%		
-%	\end{enumerate}
-%	
-%\end{itemize}
-%
-\subsubsection{Task Structure}
-
-The subjects were given a skeleton class which contained the instruction for how to complete the task and a method stub called \texttt{getRegexPattern} for each task.
-They are asked to fill the body of the \texttt{getRegexPattern} method.
-The \texttt{Task1.java} - \texttt{Task9.java} files were listed in the Eclipse project explorer so that the subjects could easily open and reopen the tasks in any order that they wanted.
-
-%\begin{figure}[h]%
-%\includegraphics[width=\columnwidth]{Task1.png}%
-%\caption{The initial content of \texttt{Task1.java} file}%
-%\label{task1}%
-%\end{figure}
-
-\section{User Study Results}
-After reading the prompt for the first task, all subjects began by searching for and reviewing documentation related to regular expressions.
-In all but one of these cases, the subject looked at the API documentation for the Java Pattern class during this initial review.
-The remaining subject, a member of the control group, referred to quick reference documentation provided by an external tool.
-In all cases, the documentation was left accessible in a browser window throughout the study, often displayed simultaneously alongside the coding window due to the large screen available for the study.
-
-\subsection{Control}
-In our previous online survey, we had found that no single strategy for writing regular expressions dominated the others.
-About 40\% of participants indicated that they would use an external tool, 30\% indicated a preference for writing a separate test script, and about 13\% of participants chose the guess-and-check strategy.
-Although our sample size is too small to make a quantitative comparison, we did observe each of these strategies in the control group.
-
-One subject began by using an external tool called regexpal.com.
-After writing a full regular expression and attempting to test it using the tool, the subject became dissatisfied with it and switched to a different tool, regextester.com.
-This tool too appeared to be unsatisfactory, as all subsequent tasks were performed without the aid of external tools or tests of any kind.
-
-The other subjects chose to write their regular expressions directly within Eclipse from the beginning.
-One of these subjects never compiled or checked the accuracy of the regular expressions beyond making sure Eclipse errors were addressed, while the other two attempted to write Java stubs within the test files to test the regular expression that they had written.
-They experienced considerable difficulty with this task, with one subject taking over 10 minutes to write the test code.
-The code could only check one example at a time and the subject used it to check a single positive example per task.
-
-In designing the regular expression palette, we had hypothesized that users would experience difficulties with two particular aspects of the Java Pattern API: that it used a factory pattern for instantiation, as opposed to the standard instantiation construct (new), and that it required special care with escape sequences -- escape sequences in the pattern must themselves be escaped, because the regular expression is written as a Java string literal.
-We observed difficulties with both of these issues in the control group.
-
-One of the four control subjects tried the \verb|new Pattern| construct.
-An Eclipse error alerted the subject to the problem shortly thereafter.
-After referring to an example in the inline API documentation for the class, the subject was able to correct this error.
-The remaining subjects all noticed this example beforehand, and were able to avoid this problem.
-
-Three of the four control subjects experienced significant difficulties associated with escape sequences.
-One subject recognized the error fairly quickly after Eclipse complained that the escape sequences used in the pattern were invalid (they were, in fact, valid escape sequences for regular expression patterns, but not for string literals.) The subject continued to miss escape sequences occasionally throughout the remainder of the study, but was able to fix them quickly after Eclipse alerted the subject to the error.
-For the other two subjects, the problem was more severe.
-In both cases, they noticed the error that Eclipse gave, but thought it indicated a problem with their pattern itself.
-One subject thought that the uses of symbolic escape sequences like \verb|\(| were incorrect.
-He decided to replace these with ASCII escape sequences (\verb|\050|) after looking up an ASCII conversion table.
-This was an unnecessary and overly complex solution to the problem.
-The other subject thought that the problem was in his use of the whitespace escape sequence, \verb|\s|.
-Thus, he  spent several minutes looking up how to match whitespace correctly.
-In doing so, he found a description of the double-escape problem in Java and was able to fix the problem after that.
-\subsection{Treatment}
-One subject was already familiar with an external tool, regexpal.com (also used by a subject in the control group), and used it variously throughout the task.
-In most cases, the subject used the external tool's quick reference documentation while using our palette for authoring and testing.
-At other times, the subject used the external tool itself.
-The subject indicated that in some cases, he was not sure that our palette was free of bugs (when something unexpectedly matched or did not match\footnote{We did not observe any actual errors related to regex matching in our tool.}), and also because the external tool provided syntax and substring highlighting, a useful feature for complex regular expressions.
-When done with the external tool, the subject pasted the pattern into our tool to generate the appropriate code.
-As such, the subject had no difficulties with escaping or factory pattern instantiation.
-
-The other two subjects in the treatment group did not use any external tools.
-Both subjects decided not to use our palette for the first task (likely because they forgot that it was available, since they spent some time looking up API documentation after our initial demo with the color palette.) They both recognized the need for factory pattern instantiation and also both had an initial error related to escaping that they were able to resolve before moving on.
-One of these subjects also began to write tests within a \verb|main| method.
-
-Beginning with the second task, both subjects remembered that a palette may be available and were able to invoke it correctly.
-They all recognized that the primary input box was where the regular expression should be entered and that the two other input boxes were for positive and negative tests, respectively.
-Two of the subjects did not initially realize that multiple tests could be entered, and that entering even a single test required pressing the \verb|Enter| key.
-One subject never fully understood this, relating after the study that he thought that the palette would notify him when he was done if the example that he had entered was not matched by the pattern.
-The other subject was able to use the test case input boxes correctly after initial experimentation.
-This was perhaps due to the wording of the prompts under these entry boxes: ``enter \textbf{a} [positive \verb1|1 negative] test case above''.
-This problem was corrected in a version of the palette developed following this study.
-
-Two of the subjects used the reinvocation feature of the palette, but neither highlighted the test cases, meaning that they had to enter new test cases every time they reinvoked the palette.
-This may have been due to the fact that our example with the Color palette involved only a single line of text (unlike the example in Figure 1).
-The third subject never used this feature.
-
-Two of the subjects expressed confusion about the meaning of the green and red backgrounds on the test cases.
-Although we provided a key saying that green meant that ``pattern matches string'', the headings of the two columns: ``[should \verb1|1 should NOT] match:'', may have caused confusion.
-We had changed the meaning of these colors due to feedback from our initial presentation of the tool in class, so it is clear that regardless of the interpretation given, some subjects were confused.
-
-Despite these difficulties, however, the basic functionality of the palette seemed to help all of the subjects.
-None struggled with issues related to escaping and instantiation, for example.
-Although the sample size is not large enough to make conclusive or quantitative judgements, it was observed that the treatment group completed more of the tasks than the test group on average (7 tasks for the treatment group vs.
-6 for the control group).
-All of the subjects in the treatment group, as well as members of the control group who were shown the palette after the study, indicated that they felt that the palette was helpful, and several provided specific suggestions for improvements.
-%\section{Future Work}
-%
-%\subsection{Palette Design Improvements}
-%The observations above suggest several potential improvements for the regular expression palette.
-Some of the problems appeared to be due to the subject's lack of familiarity with the Graphite system.
-In particular, subjects did not highlight the full text needed to reinvoke the palette, likely due to the the limited training protocol.
-By showing only a single, simple example, the subjects were not able to generalize the reversibility mechanism to multiline completions.
-If the training period had included such an example, we suspect that this would not have been a problem.
-%
-%In other cases, more explicit labels in the palette may have helped.
-The subjects began by entering only a single example because the label on the palette used the singular article, ``a''.
-We suspect that we could have avoided this problem if the label was more explicit: ``Create positive examples by entering them above and pressing ENTER''.
-Confusion over the meaning of the green and red backgrounds may also have been avoided by using a clearer label for the key, accompanied by a label explaining the meaning of each column more explicitly.
-%
-%In conjunction with these improvements, it would be useful to have an explicit ``Help'' button that allowed users to view documentation or a training video for the palette.
-On first invocation, it may also have been helpful to invoke this video automatically (though this would require tool support for associating data with the user.)
-%
-%Several subjects suggested new features for the regular expression palette, including syntax highlighting, substring highlighting and an explicit option to match the whole string, rather than any part of the string (a problem encountered by several subjects in both groups).
-They also suggested including inline quick reference documentation and regex snippets directly within the palette.
-Other people (not a part of the study) have suggested other features, including a tool for building regular expressions directly from examples, and a tool for generating examples automatically from regular expressions.
-Examples of such tools exist online.
-%
-%\subsection{Study Design Improvements}
-%
-%Some of the limitations in our study design, combined with the small size of our subject pool, made any sort of quantitative analysis difficult.
-In this section, we suggest some possible improvements.
-%
-%	\paragraph{Task Difficulty} Subjects were sometimes ``stumped'' by some of the more difficult tasks.
-It would perhaps be helpful to remove some of these excessively difficult tasks.
-For example, validating US phone numbers and dates turned out to be much more difficult than the others.
-%	
-%	\paragraph{Giving a time limit for each task} Because we only imposed a time limit for the whole experiment, it was difficult to compare performance on an individual task, in terms of time spent.
-However, imposing a per-task time limit may introduce concerns about the learning effect -- most subjects spent a significant amount of time looking up documentation in the early part of the study.
-One possibility may be to introduce dummy tasks at the beginning of the experiment.
-%	
-%	\paragraph{Using other types of Graphite palettes} This study does not evaluate the whole Graphite plug-in architecture, and many of the issues may have been due to the regular expression task and palette specifically.
-Designing a study to effectively evaluate the design as a whole would likely require building several more high-quality palettes.
-%	
-%	[Evaluating performance] It is difficult to measure accuracy of provided regular expressions due to their inherent complexity.
-We could not run an automated test suite on the subject responses because many subjects had syntax errors or other problems in their final submission that would have caused trivial failures.
-	Poor-quality regular expressions written in a relatively short amount of time may also be hard to compare to high-quality regular expressions written in a longer amount of time.
-%		
-\section{Related Work}
-In addition to the references provided in the introduction, some other concepts are related to active code completion.
-
-\subsection{Active Libraries}
-We named this technique {\it active code completion} because of its relation to the general concept of {\it active libraries} \cite{activelibraries}.
-Active libraries are libraries that contain program logic that is invoked at either compile-time or, here, design-time.
-
-\subsection{Visual Languages}
-Because active code completion involves graphical user interface elements but ultimately generates textual source code representations, it can be considered a hybrid approach that borrows interaction techniques from visual languages while remaining compatible with conventional programming languages.
-This hybrid approach may help address some of the usability challenges previously associated with visual languages (cf.
-\cite{oai:CiteSeerXPSU:10.1.1.110.3146}).
-
-Editing environments like Barista \cite{Ko06} has previously been developed that merge concepts from both conventional text based and structured editors.
-Barista allows developers to use alternative code representations.
-It differs from Graphite in that these representations are built directly into the editor (although adding new representations is possible), and that Barista requires the users to use its own model-view-controller based language, which is likely to be unfamiliar to many users.
-In future work, we hope to explore an extensible, keyboard-driven, code-generation based approach that leverages structured code representations to eliminate the difficulties associated with reinvocation and maintaining palette state described in this paper.
-
-\subsection{Specific IDE Features}
-There exist some IDE features that have been specifically designed for certain types.
-For example, CodeRush has a ``show color dialog'' \cite{CodeRush} which allows the developers to launch a color picker directly from the code editor.
-Resharper \cite{Resharper} also has a color assistance tool which previews the actual colors right beside the code completion menu items.
-However, these IDE specific features are hard-coded -- user-defined types cannot provide similar functionality.
-Recent versions of Visual Studio have started to explore user-defined palettes for widgets in the windowing system.
-
-\section{Conclusion}
-Motivated by evidence that integrating highly-specialized tools directly into a developer's workflow is useful, we have developed the concept of active code completion as a generalization of conventional code completion.
-We validated the usefulness and generated a number of use cases and design constraints for this concept with an extensive survey of professional developers.
-Based on these findings, we developed an active code completion architecture that makes several novel design decisions to ease development, deployment and discovery of user-defined palettes.
-We created a palette for regular expression tasks and validated its usefulness with a user study.
-We claim that this provides evidence for the more general claim that integrating palettes into a code editor is useful, and that active code completion systems like Graphite considerably ease this process.
 
@@
 % use section* for acknowledgement
-\section*{Acknowledgment}
+%\section*{Acknowledgment}
+
+%The authors would like to thank...
 
@@
 % http://www.michaelshell.org/tex/ieeetran/bibtex/
-%\bibliographystyle{IEEEtran}
+\bibliographystyle{IEEEtran}
 % argument is your BibTeX string definitions and bibliography database(s)
-%\bibliography{IEEEabrv,../bib/paper}
-%
-% <OR> manually copy in the resultant .bbl file
-% set second argument of \begin to the number of references
-% (used to reserve space for the reference number labels box)
-\bibliographystyle{IEEEtran}
-%\bibliography{graphite-icse12}
-\bibliography{../research}
+\bibliography{vlhcc11}
 

### 2012-03-18 vlhcc11/survey-report.tex
Initial paper move.

--- 2012-03-18 vlhcc11/survey-part.tex
@@
 \usepackage{rotating}
-\usepackage{url}
 
@@
 % can use linebreaks \\ within to get better formatting as desired
-\title{Active Code Completion}
+\title{Usability Criteria for Graphical Code Completion}
 
@@
 % affiliations
-\author{\IEEEauthorblockN{Cyrus Omar, YoungSeok Yoon, Thomas D. LaToza, Brad A. Myers}
-\IEEEauthorblockA{School of Computer Science\\
+\author{\IEEEauthorblockN{Cyrus Omar}
+\IEEEauthorblockA{Computer Science Department\\
 Carnegie Mellon University\\
-Pittsburgh, PA, USA\\
-\{comar,youngseok,tlatoza,bam\}@cs.cmu.edu}
-\texttt{\url{https://github.com/cyrus-/graphite}}
+Pittsburgh, PA 15213\\
+cyrus@cmu.edu}
+\and
+\IEEEauthorblockN{YoungSeok Yoon}
+\IEEEauthorblockA{Institute for Software Research\\
+Carnegie Mellon University\\
+Pittsburgh, PA 15213\\
+youngseok@cs.cmu.edu}
 }
@@
 % make the title area
-% \maketitle
+\maketitle
 
+%\begin{abstract}
+%\boldmath
+%\end{abstract}
 % IEEEtran.cls defaults to using nonbold math in the Abstract.
@@
 % creates the second title. It will be ignored for other modes.
-% \IEEEpeerreviewmaketitle
+\IEEEpeerreviewmaketitle
 
-\section{Developer Survey}
+\section{Introduction}
+Many source code editors today feature interactive code completion.
+By invoking a keyboard shortcut, developers can summon a floating palette containing a list of contextually relevant variables, fields, methods, classes and code snippets.
+An adjacent floating window displays documentation for the currently highlighted item.
+This interface helps developers explore APIs and avoid basic spelling and logic errors [?], and empirical data indicates that professional developers invoke interactive code completion frequently \cite{murphy_how_2006}.
 
-Prior to implementation, we conducted a large online survey of professional software developers in order to extract criteria that govern whether our system would be useful to them.
-More specifically, we sought to address the following questions:
+A number of new to the code completion list have been proposed in the literature.
+These methods combine information about the program context with make use of usage history [?], examples extracted from code repositories [?, ?] and crowdsourced information [?]
 
-\begin{itemize}
-\item What are specific use cases for active code completion in a professional setting? 
-\item What are some functional criteria that predict whether a class would benefit from active code completion?
-\item What are some usability criteria that should govern palette interface designs?
-\item Which capabilities must a active code completion architecture support to enable these use cases and designs?
-\end{itemize}
+Calcite, reordering, crowdsourced code completion papers.
 
-% In order to convey the active code completion concept clearly, we came up with three different use cases where we belive the active code completion technique will benefit developers frequently.
-For each use case, we designed three mock-up screenshots and showed them in sequence: before invoking the palette, during the interaction with the palette, and after the palette is closed and code is generated.
-By doing this, the respondents could understand the concept precisely and give useful feedback relevant to the above questions.
+Our proposal: graphical code completion. Inversion of control: look at class. Also possible to look at external database. Visual languages have shown this might be useful in certain circumstances (?).
 
-We posted a link to our survey on a popular  collaborative filtering website \cite{mehta_robust_2007} called {\tt /r/programming} on reddit.com\footnote{\texttt{http://www.reddit.com/r/programming}}.
-At the time of our study, the website listed approximately 340,000, though members were not required to registered to read the content,  and contained content that could be considered relevant to our intended audience.
-An additional 22 participants were recruited using a mass email to local computer science graduate students.
-The recruitment materials in both cases stated that we were seeking developers ``familiar with an object-oriented programming language like Java, C\# or Visual Basic and an integrated development environment like Eclipse or Visual Studio''.
+Consistent with best practices, we begin by seeking the usability criteria that govern whether such a system would be useful to professional developers.
+We conducted a large survey of these developers and extracted several criteria that both constrain the overall system architecture and inform the design (or caution against designing) palettes for different kinds of classes.
+
+%\section{Tool}
+%Described earlier.
+
+\section{Methods}
+Our subject pool consisted primarily of visitors to a large collaborative filtering website dedicated to programming called {\tt /r/programming} on reddit.com\footnote{http://www.reddit.com/r/programming}.
+At the time of our study, there were approximately 340,000 registered members.
+A small number of participants were also recruited using a mass email to local computer science graduate students.
+The recruitment materials in both cases stated that we were seeking developers ''familiar with an object-oriented programming language like Java, C\# or Visual Basic and an integrated development environment like Eclipse or Visual Studio''.
 
 Participants were asked to take an online survey that would take approximately 20 minutes to complete.
-Of the 696 people who started the survey by answering the first question, 473 participants completed it.
-Of these, 16 participants were computer science graduate students at Carnegie Mellon and 457 were from \texttt{/r/programming}.
+Of the 696 participants who started the survey, 475 participants finished it.
+Of these, 457 participants came from /r/programming and 18 from the graduate student mailing list.
+All quantitative metrics used in this study are based only on these participants.
 
-Their responses revealed a number of interesting use cases and non-trivial design constraints for active code completion systems.
-Many of these findings can also be generalized to constrain the design of visual programming languages, particularly those that aim be useful to professional developers, rather than novice programmers (e.g., Barista \cite{ko_barista:_2006}).
-We summarize the survey questions and the results in the following subsections.
-% In fact, active code completion can be considered a hybrid approach that borrows interaction techniques from visual languages while remaining compatible with conventional representations of program logic.
-This may help address some of the usability challenges previously associated with visual languages (cf.
-[?]).
+\subsection{Familiarity with Languages and Editors}
 
-\subsection{Familiarity with Programming Languages and Editors}
-
-We first asked their level of familiarity with several programming languages, on a five-point Likert scale ranging from ``None'' to ``Expert''.
-61.1\% of the respondents expressed themselves as experts on at least one language, and 35.7\% were very familiar with at least one language.
+We began by asking participants to rate their level of familiarity with several programming languages.
+Figure 1 summarizes these ratings.
 On average, participants rated themselves as very familiar with Java, C, C++ and JavaScript, familiar with C\#, Python and PHP and somewhat familiar with Visual Basic and Perl.
 
-We also asked participants to select the integrated development environments (IDEs) and code editors that they were familiar with.
+We also asked participants to select the integrated development environments and code editors that they were familiar with.
 The Eclipse IDE was familiar to 87.1\% of participants.
-This was followed by Visual Studio at 66.0\%, Vi/Vim at 53.7\%, Netbeans at 37.7\%, Emacs at 24.8\% and IntelliJ IDEA at 16.4\%.
+This was followed by Visual Studio with 66.0\% of participants, Vi/Vim with 53.7\%, Netbeans with 37.7\%, Emacs with 24.8\% and IntelliJ IDEA with 16.4\%.
 Participants could also enter ``other'' choices.
-A number of editors and IDEs were entered, including Xcode, Textmate and Notepad++.
-None of these were dominant.
+A number of editors and IDEs were entered, including Xcode, Textmate, Notepad++ and others.
+%\begin{figure}
+%  \includegraphics[width=3.5in,height=2in]{experience.png}
+%  \caption{Participant familiarity with various programming languages. Starting at bottom: None, Somewhat familiar, Familiar, Very Familiar, Expert.}
+%\end{figure}
 
-\subsection{Mock-ups}
+\subsection{Mockups}
+\begin{figure*}
+  \includegraphics[scale=.6]{mockup-palette.png}
+  \caption{The regular expression palette, shown in Eclipse.}
+\end{figure*}
+Next, we presented participants with a series of mockup palettes corresponding to a Color class, a regular expression Pattern class (Figure 2) and a SQL query ResultSet class.
+In addition to a screenshot of the palette itself, participants were shown mockups demonstrating how a user would invoke the palette, and a mockup showing the code that would be inserted once a selection had been made.
 
-\begin{figure}
-  \includegraphics[width=\linewidth]{mockup-palette-small.png}
-  \caption{An example mock-up palette for regular expressions shown to the survey participants. In the survey, they were shown the surrounding Eclipse IDE together.}
-  \label{regex}
-\end{figure}
+Before presenting each palette mockup we gathered information about the user's familiarity with that topic and the strategies that they would likely use to instantiate that class.
+The Color class was presumed familiar to all participants.
+Figure 3 summarizes responses for the regular expression and SQL classes.
+The low number of participants unfamiliar with these topics again provides evidence that our participants were not generally novice developers.
 
-Next, we presented participants with a series of mock-up palettes for a Color class, a regular expression class (Fig.
-\ref{regex}), and a SQL query class.
-Participants were also shown a mock-up demonstrating how a user could invoke the palette, and a mock-up showing the code that would be inserted once a selection had been made.
+To determine the strategy typically used for instantiating the Color class, we asked the following question. The values in bold show the frequency of each response.
 
-Before presenting each mock-up, we gathered information about the user's familiarity with the class and the strategies that they would likely use to instantiate the class under normal circumstances.
-For the Color class, the majority of participants indicated that they would look in the code completions menu (58.4\%) or in the class documentation (19.0\%) for a predefined constant.
-Another 14.0\% indicated that they would use an external tool to determine the RGB values.
-Few participants were unfamiliar with regular expressions and no participants were unfamiliar with SQL queries which provides further evidence that our participants were not novice developers.
-Fig.
-\ref{strategies} summarizes their usual strategies for instantiating regular expressions and SQL queries.
+\begin{quote}
+A user interface designer asks you to change the background color of a window to ``navy blue'' when the user presses a button.
+You have determined that you need to instantiate the 'Color' class to do so.
+\\\\
+Which of the following strategies would you use first to create an instance of the 'Color' class representing navy blue?\\
+
+\begin{enumerate}[(a)]
+\item I would use the IDE's content assist menu (see note below) to look for a predefined constant representing navy blue. (\textbf{58.4\%})
+\item I would look up the documentation of the Color class to see if there is a predefined constant representing navy blue. (\textbf{19.0\%})
+\item I would use an external tool (e.g. Photoshop or some other image editor) to determine the RGB values for navy blue. (\textbf{14.0\%})
+\item I would guess at the RGB values for navy blue, then check whether my guess was correct by executing the application. (\textbf{5.3\%})
+\item Other (please specify) (\textbf{3.3\%})\\
+\end{enumerate}
+\end{quote}
+
+Figure 4 summarizes the responses to a more general question asked about regular expressions and SQL queries.
 
@@
 \begin{tabular}{lccc}
- & Regular Expressions &  & SQL\\
+ & \textbf{Regular Expressions} &  & \textbf{SQL}\\
  \hline
-Separate test script & 29.6\% &  & 15.4\%\\
-Guess and check & 14.0\% &  & 16.1\%\\
-External tool & \textbf{37.9\%} &  & \textbf{58.6\%}\\
-Search for examples & 12.3\% & & 5.1\% \\
-Other & 6.2\% & & 4.9\% \\
+Never used & 4.8\% &  & 0.0\%\\
+Use infrequently & 46.7\% &  & 37.4\%\\
+Use frequently & \textbf{48.4\%} &  & \textbf{62.6\%}\\
+\hline
+\end{tabular}
+\caption{Participant's experience with regular expressions and SQL.}
+\end{figure}
+
+\begin{figure}
+\begin{tabular}{lccc}
+ & \textbf{Regular Expressions} &  & \textbf{SQL}\\
+ \hline
+Separate test script & 29.7\% &  & 15.5\%\\
+Guess and check & 13.4\% &  & 16.0\%\\
+External tool & \textbf{38.5\%} &  & \textbf{58.9\%}\\
+Search for examples & 12.1\% & & 4.8\% \\
+Other & 6.2\% & & 4.8\% \\
 \hline
@@
 \caption{Typical strategies for regular expressions and SQL queries.}
-\label{strategies}
 \end{figure}
+
+%For each of these classes,
+
+%\subsubsection{Color}
+%We began with a class representing colors. We asked participants to consider the following scenario:
 
@@
 \end{tabular}
-\caption{The distribution of responses to the question: ``Consider situations where you need to instantiate the [specified] class.
+\caption{For each of the classes considered, participants were shown mockups of a code completion palette designed for instantiating that class, as well as the source code generated after parameters were entered.
+This table shows the distribution of responses to the question: ``Consider situations where you need to instantiate the [specified] class.
 What portion of the time, in these situations, do you think you would use this feature?''}
-\label{frequencies}
+\end{figure}
+%
+%\begin{figure}
+%  \includegraphics[scale=.6]{regex-insert.png}
+%  \caption{Test}
+%\end{figure}
+%
+%
+
+%Then we asked:
+%\begin{quote}
+%\textbf{Q2} Consider situations where you need to instantiate the 'Color' class. What portion of the time, in these situations, do you think you would use this feature?
+%\begin{enumerate}[(a)]
+%\item Nearly every time (\textbf{9.6\%})
+%\item Most of the time (\textbf{22.1\%})
+%\item Some of the time (\textbf{32.4\%})
+%\item Rarely (\textbf{28.2\%})
+%\item Never (\textbf{7.7\%})\\
+%\end{enumerate}
+%\end{quote}
+%
+
+After presenting each palette, we asked users to rate the usefulness of each palette.
+Responses are summarized in Figure 5.
+We then allowed participants to provide qualifications and open-ended feedback by asking the following question: ``If you would like to qualify or elaborate on your answer to the previous question you may do so below.
+You may also suggest improvements for this palette.'' This information is analyzed in the remainder of the paper.
+
+%\begin{quote}
+%\textbf{Q4} How would you describe your level of familiarity with (regular expressions, SQL)?\\
+%\\
+%\begin{enumerate}[(a)]
+%\item I do not know what [it is].	 (\textbf{0.7\%}, \textbf{0.0\%})
+%\item I know what regular expressions are but I have not used them in practice. (\textbf{4.4\%, 0.0\%})
+%\item I use regular expressions infrequently.	 (\textbf{46.0\%, 35.9\%})
+%\item I use regular expressions frequently. (\textbf{49.0\%, 64.1\%})
+%\end{enumerate}
+%\end{quote}
+
+%\begin{quote}
+%\textbf{Q5} Consider situations where you need to write a regular expression to extract information from a string or collection of strings. What strategy would you most likely use first?
+
+%\begin{enumerate}[(a)]
+%\item I would write a separate test script to help come up with an appropriate regular expression. (\textbf{29.7\%})
+%\item I would guess and check regular expressions by running my application. (\textbf{13.4\%})
+%\item I would launch an external application or web tool to help me test regular expressions interactively. (\textbf{38.5\%})
+%\item I would search the web for a sample regular expression suitable for my task. (\textbf{12.1\%})
+%\item Other (please specify) (\textbf{6.2\%})
+%\end{enumerate}
+%\end{quote}
+
+%\begin{quote}
+%\textbf{Q6} Consider situations where you need to create a regular expression. What portion of the time, in these situations, do you think you would use this feature?
+%\begin{enumerate}[(a)]
+%\item Nearly every time (\textbf{36.6\%})
+%\item Most of the time (\textbf{29.5\%})
+%\item Some of the time (\textbf{21.8\%})
+%\item Rarely (\textbf{7.3\%})
+%\item Never (\textbf{4.8\%})\\
+%\end{enumerate}
+%\end{quote}
+
+%\begin{quote}
+%\textbf{Q7} How would you describe your level of familiarity with SQL?
+
+%\begin{enumerate}[(a)]
+%\item I do not know what regular expressions are.	 (\textbf{0.0\%})
+%\item I know what SQL is I have not used it in practice. (\textbf{0.0\%})
+%\item I use SQL infrequently.	 (\textbf{35.9\%})
+%\item I use SQL frequently. (\textbf{64.1\%})\\
+%\end{enumerate}
+%\end{quote}
+
+%\begin{quote}
+%\textbf{Q8} Consider situations where you need to write a regular expression to extract information from a string or collection of strings. What strategy would you most likely use first?
+
+%\begin{enumerate}[(a)]
+%\item I would write a separate test script to help come up with an appropriate SQL query.	
+ %(\textbf{15.5\%})
+%\item I would guess and check by running my application. (\textbf{16.0\%})
+%\item I would launch an external application or web tool to test my query against my database interactively.	(\textbf{58.9\%})
+%\item I would search the web for an example SQL query suitable for my task.	
+ %(\textbf{4.8\%})
+%\item Other (please specify) (\textbf{4.8\%})
+%\end{enumerate}
+%\end{quote}
+
+%Showed the SQL palette and asked:
+
+%\begin{quote}
+%\textbf{Q9} Consider situations where you need to create a SQL query. How often, in these situations, do you think you would use this feature?
+%\end{quote}
+
+\section{Design Implications}
+\begin{figure}[b]
+  \includegraphics[scale=.5]{color-palette.png}
+  \caption{test}
 \end{figure}
 
-Participants were positive about our mock-ups.
-We asked the users to rate the usefulness of each mock-up palette, and their responses are summarized in Fig.
-\ref{frequencies}.
-For each palette, more than half of the respondents expressed that they would use it at least some of the time.
-Especially, people liked the regular expression palette a lot.
-
-\section{Design Implications}
-
-We analyzed the open-ended responses to extract design principles and constraints.
+We next analyzed the open-ended responses to extract some design principles and constraints.
 We got 193 responses for the color palette, 129 for the regular expression palette, and 142 for the SQL palette.
 Also, we got total 119 suggestions about what other types of classes could benefit from similar types of palettes.
-The extracted principles and constraints are summarized in the following subsections.
-The parenthesized number following each design constraint indicates how many people commented on that issue.
+The design can be divided into two main categories: system design constraints, and general palette design considerations.
+System design constraints are the criteria that should be taken into account in the design of the overall system architecture, and the general palette design considerations are about how to make more usable palettes.
 
@@
 
-System design constraints are the criteria that should be taken into account in the design of the overall system architecture.
+\subsubsection{Handling Separation of Concerns}
 
-\subsubsection{Handling Separation of Concerns (183?)}
+Professional developers were very wary of including constant data into the source code directly.
+Many responses expressed general sentiment or specific preference for project-wide color theme or external resource file.
+Also, many people stated that these tools might not be useful since they usually read an external resource file, or sometimes they explicitly suggested to add capability to handle the external resource files directly from the palette.
 
-Participants were very wary of including constant data into the source code directly.
-Many responses expressed general sentiment or specific preference for project-wide color theme or external resource file.
-Sometimes they explicitly suggested to add capability to handle the external resource files directly from the palette.
+Examples of feedbacks in this category include:
+
+\begin{itemize}
+	\item The use of magic numbers in the source code is a bad design (54)
+	\item Color picking is the `designer's job' (33)
+	\item Would only use to define named constants (17)
+	\item Would rather use color scheme / theme (3)
+	\item Would use Object-Relational Mapping (ORM) / Language INtegrated Query (LINQ) instead of raw SQL query string (27 of 142)
+	\item These palettes might be useful for prototyping, but not for real production (14)
+	\item Instead of including the test cases into comments, they should be stored as unit tests (35 of 143)
+\end{itemize}
+
 Interestingly, most people seemed to consider regular expressions as part of the program logic.
-There were few people who complained about including the regular expression pattern string in the source code as opposed to the other two classes.
+There were few people who complained about including the regular expression pattern string in the source code.
 
-%Examples of feedbacks in this category include:
-%
-%\begin{itemize}
-%	\item The use of magic numbers in the source code is a bad design (54)
-%	\item Color picking is the `designer's job' (33)
-%	\item Would only use to define named constants (17)
-%	\item Would rather use color scheme / theme (3)
-%	\item Would use Object-Relational Mapping (ORM) / Language INtegrated Query (LINQ) instead of raw SQL query string (27 of 142)
-%	\item These palettes might be useful for prototyping, but not for real production (14)
-%	\item Instead of including the test cases into comments, they should be stored as unit tests (35 of 143)
-%\end{itemize}
+\subsubsection{Reversibility}
 
-\subsubsection{Reversibility (19?)}
+19 participants expressed that it should be able to bring the palettes back when they want to modify the parameters after they used the palettes.
 
-Some participants expressed that it should be able to bring the palettes back when they want to modify the parameters after they used the palettes.
+\subsubsection{Palette Settings and State}
 
-\subsubsection{Palette Settings and State (41?)}
+Many participants wanted to be able to configure the palettes or have the palettes maintain state even after the palettes are hidden. The followings are the comments that are related to this category.
 
-Many participants wanted to be able to configure the palettes or have the palettes maintain state even after the palettes are hidden.
-% The followings are the comments that are related to this category.
+\begin{itemize}
+	\item Want recently used regular expressions, and control over the comments (12)
+	\item Persistent database connection information is needed / want to manage connection pools (9)
+	\item Want recent colors, capability of setting favorite colors (20)
+\end{itemize}
 
-%\begin{itemize}
-%	\item Want recently used regular expressions, and control over the comments (12)
-%	\item Persistent database connection information is needed / want to manage connection pools (9)
-%	\item Want recent colors, capability of setting favorite colors (20)
-%\end{itemize}
-
-\subsubsection{Interaction with Code Context (13?)}
+\subsubsection{Interaction with Code Context}
 
 Some people wanted to interact with the code context.
-For example, people expressed that it would be great if the SQL query palette was capable of dynamically creating SQL query string by combining multiple string variables, or by assigning variables for the parameters.
+For example, 13 people expressed that it would be great if the SQL query palette was capable of dynamically creating SQL query string by combining multiple string variables, or by assigning variables for the parameters.
 
-\subsubsection{Performance (?)}
+\subsubsection{Performance}
 
@@
 
-\subsubsection{IDE Independence (?)}
+\subsubsection{IDE Independence}
 
@@
 
+%\subsubsection{Composing palette logic}
+
 \subsection{General Palette Design Considerations}
-
-General palette design considerations are about how to make more usable palettes.
-
 \subsubsection{Simplicity vs. Capability}
@@
 
-\subsubsection{Keyboard Navigability (12)}
+\subsubsection{Keyboard Navigability}
 
@@
 
-\subsection{Suggestions from the Participants}
-At the end of the survey, we solicited suggestions from the participants about what types of classes could benefit from the active code completion palettes.
-119 participants suggested various types of classes and we classified the suggestions into several categories.
+\subsubsection{Complexity}
 
-\subsubsection{Alternative/Tricky/Literal Syntax (16)}
-These are the classes of which syntax is tricky, or complex.
-Even for a basic string class, people wanted to input multiline string or unicode string in more convenient way with automatic escaping.
-We address this issue in more detail in Section \#.
-%
-%\begin{itemize}
-%	\item Example classes
-%	
-%	\begin{itemize}
-%		\item String
-%		\item Collections (e.g., dictionary)
-%		\item Vector, Matrix
-%		\item Embedded languages
-%		\item URLs, Paths
-%	\end{itemize}
-%	
-%	\item Expected features
-%	
-%	\begin{itemize}
-%		\item Proper escaping
-%		\item Syntax highlighting
-%		\item Auto-completion
-%	\end{itemize}
-%\end{itemize}
+%\subsubsection{Side Effects}
 
-\subsubsection{Unclear Parameter Implications (11)}
+\section{Suggestions from the Participants}
+We classified the suggestions into several categories. The number in the parentheses indicate the number of participants who suggested the classes which fall into the category.
+
+\subsection{Alternative/Tricky/Literal Syntax (16)}
+These are the classes of which syntax is tricky, or complex. Even for a basic string class, people wanted to input multiline string or unicode string in more convenient way with automatic escaping.
+
+\begin{itemize}
+	\item Example classes
+
+	\begin{itemize}
+		\item String
+		\item Collections (e.g., dictionary)
+		\item Vector, Matrix
+		\item Embedded languages
+		\item URLs, Paths
+	\end{itemize}
+
+	\item Expected features
+
+	\begin{itemize}
+		\item Proper escaping
+		\item Syntax highlighting
+		\item Auto-completion
+	\end{itemize}
+\end{itemize}
+
+\subsection{Unclear Parameter Implications (11)}
 The classes in this category has some parameters in it, and it is not easy to predict the results of the parameter modifications.
 Instead of running the whole application to see the result, the palette can serve as a preview panel so that the developers may check the results directly.
-Also, the palette can be a control panel which facilitates modifying the parameters as in Juxtapose [?].
+Also, the palette can be a control panel which facilitates modifying the parameters.
+\begin{itemize}
+	\item Example classes
 
-%\begin{itemize}
-%	\item Example classes
-%	
-%	\begin{itemize}
-%		\item Audio tweaks (modifying pitch, volume, etc.)
-%		\item 3D transformation matrices
-%		\item number / string / date formatting
-%		\item message box / input box (e.g., \texttt{JOptionPane})
-%	\end{itemize}
-%\end{itemize}
+	\begin{itemize}
+		\item Audio tweaks (modifying pitch, volume, etc.)
+		\item 3D transformation matrices
+		\item number / string / date formatting
+		\item message box / input box (e.g., \texttt{JOptionPane})
+	\end{itemize}
+\end{itemize}
 
-\subsubsection{Query Languages (17)}
-Similar to the regular expression and SQL palette we presented, the developers also wanted to check if the other types of query strings (e.g., XPath / XQuery) are correct or not by checking the results directly on the palette.
+\subsection{Query Languages (17)}
+Similar to the unclear parameter implications category, the developers wanted to check if the query string is correct or not by checking the results directly on the palette.
 
-%\begin{itemize}
-%	\item Example classes
-%	
-%	\begin{itemize}
-%		\item Regular expression
-%		\item SQL query
-%		\item XPath / XQuery
-%	\end{itemize}
-%\end{itemize}
+\begin{itemize}
+	\item Example classes
 
-\subsubsection{Graphical Elements (27)}
+	\begin{itemize}
+		\item Regular expression
+		\item SQL query
+		\item XPath / XQuery
+	\end{itemize}
+\end{itemize}
+
+\subsection{Graphical Elements (27)}
 The other popular category was graphical elements.
@@
 
-%\begin{itemize}
-%	\item Example classes
-%	
-%	\begin{itemize}
-%		\item Color, Shape, Brush, Font, etc.
-%		\item JFrame, Layouts, any swing components
-%	\end{itemize}
-%\end{itemize}
+\begin{itemize}
+	\item Example classes
+
+	\begin{itemize}
+		\item Color, Shape, Brush, Font, etc.
+		\item JFrame, Layouts, any swing components
+	\end{itemize}
+\end{itemize}
 
@@
 For example, in order to read a text file in Java, the developer might want to use \texttt{BufferedReader} class.
-It is error prone to use this class because it requires try/catch block, and also one must close the file after reading it.
+The instantiation code might look as following:
+\begin{verbatim}
+BufferedReader reader = null;
+try {
+    reader = new BufferedReader(
+        new FileReader(filePath));
+} catch (IOException e) {
+    e.printStackTrace();
+}
+// use reader here
+reader.close();
+\end{verbatim}
 By using a palette to choose a file or choose a variable which contains the file path, the developers could easily instantiate these objects.
-In the same way, it can alleviate the factory pattern usability problem \cite{ellis_factory_2007}.
+In the same way, it can alleviate the factory pattern usability problem.
 As long as the developers remember which class to use, they will not need to remember how to instantiate that class.
 
-%\begin{verbatim}
-%BufferedReader reader = null;
-%try {
-%    reader = new BufferedReader(
-%        new FileReader(filePath));
-%} catch (IOException e) {
-%    e.printStackTrace();
-%}
-%// use reader here
-%reader.close();
-%\end{verbatim}
+\begin{itemize}
+	\item Example classes
+
+	\begin{itemize}
+		\item BufferedReader
+		\item Classes using factory patterns
+		\item Database connection string builder
+	\end{itemize}
+\end{itemize}
 
@@
 
-%\begin{itemize}
-%	\item Example classes
-%	
-%	\begin{itemize}
-%		\item Keyboard Keys
-%		\item Regular expression
-%	\end{itemize}
-%\end{itemize}
+\begin{itemize}
+	\item Example classes
+
+	\begin{itemize}
+		\item Keyboard Keys
+		\item Regular expression
+	\end{itemize}
+\end{itemize}
 
@@
 If the developers are not familiar of the specific API class, the palette could tell them how to use it.
+
+%\begin{itemize}
+%\item 54 expressed a concern about encouraging inlining of constants, preferring external resource files.
+17 stated that they would use this feature only to define constants.
+33 stated that it should be the designer's job.
+%\item 11 expressed a concern about keyboard navigability
+%\item 27 expressed a concern about the complexity of the palette
+%\item 10 expressed an interest in a color picker. 7 wanted a multitabbed window with different options for picking colors.
+%\item 26 expressed a desire for color names, 10 of which wanted to ensure color constant names were used if available
+%\item 20 wanted persistent state (MRU, favorites)
+%\end{itemize}
+
+% An example of a floating figure using the graphicx package.
+% Note that \label must occur AFTER (or within) \caption.
+% For figures, \caption should occur after the \includegraphics.
+% Note that IEEEtran v1.7 and later has special internal code that
+% is designed to preserve the operation of \label within \caption
+% even when the captionsoff option is in effect. However, because
+% of issues like this, it may be the safest practice to put all your
+% \label just after \caption rather than within \caption{}.
+%
+% Reminder: the "draftcls" or "draftclsnofoot", not "draft", class
+% option should be used if it is desired that the figures are to be
+% displayed while in draft mode.
+%
+%\begin{figure}[!t]
+%\centering
+%\includegraphics[width=2.5in]{myfigure}
+% where an .eps filename suffix will be assumed under latex, 
+% and a .pdf suffix will be assumed for pdflatex; or what has been declared
+% via \DeclareGraphicsExtensions.
+%\caption{Simulation Results}
+%\label{fig_sim}
+%\end{figure}
+
+% Note that IEEE typically puts floats only at the top, even when this
+% results in a large percentage of a column being occupied by floats.
+
+% An example of a double column floating figure using two subfigures.
+% (The subfig.sty package must be loaded for this to work.)
+% The subfigure \label commands are set within each subfloat command, the
+% \label for the overall figure must come after \caption.
+% \hfil must be used as a separator to get equal spacing.
+% The subfigure.sty package works much the same way, except \subfigure is
+% used instead of \subfloat.
+%
+%\begin{figure*}[!t]
+%\centerline{\subfloat[Case I]\includegraphics[width=2.5in]{subfigcase1}%
+%\label{fig_first_case}}
+%\hfil
+%\subfloat[Case II]{\includegraphics[width=2.5in]{subfigcase2}%
+%\label{fig_second_case}}}
+%\caption{Simulation results}
+%\label{fig_sim}
+%\end{figure*}
+%
+% Note that often IEEE papers with subfigures do not employ subfigure
+% captions (using the optional argument to \subfloat), but instead will
+% reference/describe all of them (a), (b), etc., within the main caption.
+
+% An example of a floating table. Note that, for IEEE style tables, the 
+% \caption command should come BEFORE the table. Table text will default to
+% \footnotesize as IEEE normally uses this smaller font for tables.
+% The \label must come after \caption as always.
+%
+%\begin{table}[!t]
+%% increase table row spacing, adjust to taste
+%\renewcommand{\arraystretch}{1.3}
+% if using array.sty, it might be a good idea to tweak the value of
+% \extrarowheight as needed to properly center the text within the cells
+%\caption{An Example of a Table}
+%\label{table_example}
+%\centering
+%% Some packages, such as MDW tools, offer better commands for making tables
+%% than the plain LaTeX2e tabular which is used here.
+%\begin{tabular}{|c||c|}
+%\hline
+%One & Two\\
+%\hline
+%Three & Four\\
+%\hline
+%\end{tabular}
+%\end{table}
+
+% Note that IEEE does not put floats in the very first column - or typically
+% anywhere on the first page for that matter. Also, in-text middle ("here")
+% positioning is not used. Most IEEE journals/conferences use top floats
+% exclusively. Note that, LaTeX2e, unlike IEEE journals/conferences, places
+% footnotes above bottom floats. This can be corrected via the \fnbelowfloat
+% command of the stfloats package.
+
+\section{Conclusion}
+Overall, professional developers appear willing to use these for a number of use cases.
+However, the usability criteria described above impose stiff requirements on the architecture.
+Of particular concern, we found that the developers preferred help with program logic, but not the constant data.
+This indicates that another area of application may be in specialized editors for resource files.
 

### 2012-03-18 vlhcc11/vlhcc11.tex
Initial paper move.

--- 2012-03-18 vlhcc11/survey-part.tex
@@
 \{comar,youngseok,tlatoza,bam\}@cs.cmu.edu}
-\texttt{\url{https://github.com/cyrus-/graphite}}
+\texttt{\url{https://github.com/cyrus~/graphite}}
 }
@@
 % make the title area
-% \maketitle
+\maketitle
 
@@
 % creates the second title. It will be ignored for other modes.
-% \IEEEpeerreviewmaketitle
+\IEEEpeerreviewmaketitle
+
+\section{Introduction}
+
+Software developers today make heavy use of the code completion features available in modern code editors \cite{murphy_how_2006}.
+By navigating and selecting from a floating menu containing the names of variables, fields, methods, types and code snippets, a developer can avoid many common spelling and logic errors, avoid unnecessary keystrokes and explore unfamiliar APIs without leaving the editor window.
+To ensure that the items featured in this menu are relevant, the editor conducts a static analysis of the surrounding code context.
+%In many editors, an area adjacent to this menu displays documentation associated with the  currently highlighted item.
+
+Several refinements and additions to the code completion menu have been previously suggested in the literature.
+These have focused on using additional sources of information, such as databases of usage history \cite{robbes_how_2008}, examples extracted from code repositories \cite{bruch_learning_2009} and crowdsourced information \cite{mooty_calcite:_2010}, to increase the relevance and sophistication of the featured items.
+Empirical evidence presented in these studies suggests that these enhancements further improve developer productivity.
+
+In this paper, we propose a complementary technique called {\it active code completion}.
+When the developer invokes the code completion menu, the editor looks for a {\it palette definition} associated with the type of the expression being entered.
+If found, an option to use this palette is added to the code completion menu.
+When the developer selects this option, source code is not inserted immediately.
+Instead, the palette definition takes control of the code completion interface.
+The developer can then interact with this interface to provide parameters and other information related to her intent, and receive immediate feedback about the effect these choices will have on the object's behavior.
+When the developer indicates that she is satisfied with these choices, the palette generates code that is inserted at the cursor.
+
+Before designing and implementing such a system, we sought to address the following questions:
+
+\begin{itemize}
+\item What are some specific use cases for active code completion in a professional development setting? 
+\item Which functional criteria are common to types that would benefit from an associated palette?
+\item What usability criteria should inform palette interface designs in this context?
+\item Which capabilities must the active code completion architecture have to enable these use cases and designs?
+\end{itemize}
+
+To answer these questions, we began by conducting a large online survey of professional developers.
+Their responses revealed a number of interesting use cases and non-trivial design constraints for active code completion systems.
+
+Based on this information, we designed and implemented an Eclipse plug-in that provides active code completion for the Java language, and implemented some example palettes atop this system. 
+Finally, we evaluated the usefulness of one such palette, designed to assist developers as they write regular expressions, with a controlled lab study.
+The data and observations from this study provide empirical evidence in support of the view that an active code completion system would be useful to professional developers.
 
 \section{Developer Survey}
+The subject pool for the preliminary survey consisted primarily of visitors to a large programming-oriented 
+collaborative filtering forum\footnote{http://www.reddit.com/r/programming}.
+A small number of participants were also recruited using a
+mass email to local computer science graduate students.
+Participants were asked to take an online survey taking  approximately 20 minutes to complete and background information collected from the 475 participants indicated that the overwhelming majority of participants were professional developers.
 
-Prior to implementation, we conducted a large online survey of professional software developers in order to extract criteria that govern whether our system would be useful to them.
-More specifically, we sought to address the following questions:
+Participants were shown three mockup palettes, along with mockups of the invocation procedure described above, for classes representing colors, regular expressions and SQL queries.
+They were asked to indicate whether they would find these palettes useful.
+Their responses are summarized in Figure 1 and indicate that these developers found the concept potentially useful, with a particular preference for the palette shown for regular expressions.
 
-\begin{itemize}
-\item What are specific use cases for active code completion in a professional setting? 
-\item What are some functional criteria that predict whether a class would benefit from active code completion?
-\item What are some usability criteria that should govern palette interface designs?
-\item Which capabilities must a active code completion architecture support to enable these use cases and designs?
-\end{itemize}
+Participants were also asked to provide free-form comments on each palette.
+These responses revealed a number of concerns about this system.
+These include issues related to handling separation of concerns, reinvocation, palette settings and state, interactions between the palette and the code context, language and editor independence, and keyboard navigability.
+These concerns strongly influenced the design of the code completion system described in the next section.
 
-% In order to convey the active code completion concept clearly, we came up with three different use cases where we belive the active code completion technique will benefit developers frequently.
-For each use case, we designed three mock-up screenshots and showed them in sequence: before invoking the palette, during the interaction with the palette, and after the palette is closed and code is generated.
-By doing this, the respondents could understand the concept precisely and give useful feedback relevant to the above questions.
-
-We posted a link to our survey on a popular  collaborative filtering website \cite{mehta_robust_2007} called {\tt /r/programming} on reddit.com\footnote{\texttt{http://www.reddit.com/r/programming}}.
-At the time of our study, the website listed approximately 340,000, though members were not required to registered to read the content,  and contained content that could be considered relevant to our intended audience.
-An additional 22 participants were recruited using a mass email to local computer science graduate students.
-The recruitment materials in both cases stated that we were seeking developers ``familiar with an object-oriented programming language like Java, C\# or Visual Basic and an integrated development environment like Eclipse or Visual Studio''.
-
-Participants were asked to take an online survey that would take approximately 20 minutes to complete.
-Of the 696 people who started the survey by answering the first question, 473 participants completed it.
-Of these, 16 participants were computer science graduate students at Carnegie Mellon and 457 were from \texttt{/r/programming}.
-
-Their responses revealed a number of interesting use cases and non-trivial design constraints for active code completion systems.
-Many of these findings can also be generalized to constrain the design of visual programming languages, particularly those that aim be useful to professional developers, rather than novice programmers (e.g., Barista \cite{ko_barista:_2006}).
-We summarize the survey questions and the results in the following subsections.
-% In fact, active code completion can be considered a hybrid approach that borrows interaction techniques from visual languages while remaining compatible with conventional representations of program logic.
-This may help address some of the usability challenges previously associated with visual languages (cf.
-[?]).
-
-\subsection{Familiarity with Programming Languages and Editors}
-
-We first asked their level of familiarity with several programming languages, on a five-point Likert scale ranging from ``None'' to ``Expert''.
-61.1\% of the respondents expressed themselves as experts on at least one language, and 35.7\% were very familiar with at least one language.
-On average, participants rated themselves as very familiar with Java, C, C++ and JavaScript, familiar with C\#, Python and PHP and somewhat familiar with Visual Basic and Perl.
-
-We also asked participants to select the integrated development environments (IDEs) and code editors that they were familiar with.
-The Eclipse IDE was familiar to 87.1\% of participants.
-This was followed by Visual Studio at 66.0\%, Vi/Vim at 53.7\%, Netbeans at 37.7\%, Emacs at 24.8\% and IntelliJ IDEA at 16.4\%.
-Participants could also enter ``other'' choices.
-A number of editors and IDEs were entered, including Xcode, Textmate and Notepad++.
-None of these were dominant.
-
-\subsection{Mock-ups}
-
-\begin{figure}
-  \includegraphics[width=\linewidth]{mockup-palette-small.png}
-  \caption{An example mock-up palette for regular expressions shown to the survey participants. In the survey, they were shown the surrounding Eclipse IDE together.}
-  \label{regex}
-\end{figure}
-
-Next, we presented participants with a series of mock-up palettes for a Color class, a regular expression class (Fig.
-\ref{regex}), and a SQL query class.
-Participants were also shown a mock-up demonstrating how a user could invoke the palette, and a mock-up showing the code that would be inserted once a selection had been made.
-
-Before presenting each mock-up, we gathered information about the user's familiarity with the class and the strategies that they would likely use to instantiate the class under normal circumstances.
-For the Color class, the majority of participants indicated that they would look in the code completions menu (58.4\%) or in the class documentation (19.0\%) for a predefined constant.
-Another 14.0\% indicated that they would use an external tool to determine the RGB values.
-Few participants were unfamiliar with regular expressions and no participants were unfamiliar with SQL queries which provides further evidence that our participants were not novice developers.
-Fig.
-\ref{strategies} summarizes their usual strategies for instantiating regular expressions and SQL queries.
-
-\begin{figure}
-\begin{tabular}{lccc}
- & Regular Expressions &  & SQL\\
- \hline
-Separate test script & 29.6\% &  & 15.4\%\\
-Guess and check & 14.0\% &  & 16.1\%\\
-External tool & \textbf{37.9\%} &  & \textbf{58.6\%}\\
-Search for examples & 12.3\% & & 5.1\% \\
-Other & 6.2\% & & 4.9\% \\
-\hline
-\end{tabular}
-\caption{Typical strategies for regular expressions and SQL queries.}
-\label{strategies}
-\end{figure}
+Finally, we solicited suggestions for classes that may benefit from an associated palette.
+We received a number of interesting suggestions, which we broadly classified into categories.
+Examples include classes that would benefit from an alternative syntax (e.g.
+dictionaries), where the implications of a parameter choice are difficult to predict (e.g.
+3D transformation matrices), where the instantiation procedure is non-trivial (e.g.
+factory methods) and where parameters could be provided by example (e.g.
+key combinations).
+\begin{figure*}\begin{center}
+  \includegraphics[scale=.4]{regex.png}\end{center}
+  %\caption{Test}
+  \vspace{-15pt}
+  \end{figure*}
 
@@
 What portion of the time, in these situations, do you think you would use this feature?''}
-\label{frequencies}
+  \vspace{-15pt}
+
 \end{figure}
+\section{Prototype Implementation}
 
-Participants were positive about our mock-ups.
-We asked the users to rate the usefulness of each mock-up palette, and their responses are summarized in Fig.
-\ref{frequencies}.
-For each palette, more than half of the respondents expressed that they would use it at least some of the time.
-Especially, people liked the regular expression palette a lot.
+Next, we implemented an active code completion system as an Eclipse Java plug-in called \textsc{Graphite}, an acronym for \textbf{Gra}phical \textbf{P}alettes \textbf{H}elp \textbf{I}nstantiate \textbf{T}ypes in the \textbf{E}ditor.
 
-\section{Design Implications}
+\subsection{Association Model}
+Graphite provides two mechanisms to associate a palette with a class.
+The {\it annotation-based association model} allows the developer of a class to associate a palette definition with it using a Java annotation, \texttt{GraphitePalette}.
+The annotation contains the URL of the palette and metadata used to control the description of the palette in the code completion menu.
+This association model is beneficial because end-users need not realize that a palette is available for a class -- when they invoke the code completion menu, they {\it discover} that a palette exists.
+This stands in contrast to relevant external tools, which must be explicitly discovered by users.
 
-We analyzed the open-ended responses to extract design principles and constraints.
-We got 193 responses for the color palette, 129 for the regular expression palette, and 142 for the SQL palette.
-Also, we got total 119 suggestions about what other types of classes could benefit from similar types of palettes.
-The extracted principles and constraints are summarized in the following subsections.
-The parenthesized number following each design constraint indicates how many people commented on that issue.
+For classes which are not amenable to direct modification, such as those in the Java standard library, an {\it external association model} is available in the Graphite preferences pane.
+Users can explicitly associate a palette with the fully-qualified name of a class using this model.
 
-\subsection{System Design Constraints}
+\subsection{Palette Implementation Model}
 
-System design constraints are the criteria that should be taken into account in the design of the overall system architecture.
+A common concern expressed in our developer survey was that active code completion should not rely on a particular editor environment.
+To satisfy this constraint, we could not use an IDE-specific user interface framework (e.g.
+SWT and JFace for Eclipse).
+Instead, we designed Graphite so that palettes were written using HTML and Javascript.
+Our preliminary study confirmed the fact that these languages are well-known amongst Java developers.
+A small Javascript bridge API was developed to allow palettes to access the current selection in the editor (used to allow reinvocation of palettes) and to insert code at the cursor.
+The Eclipse plug-in simply displays the palette in a floating browser window, without any associated chrome, and responds to invocations of bridge functions.
 
-\subsubsection{Handling Separation of Concerns (183?)}
+This strategy could be straightforwardly replicated in other editor environments, without requiring that palette definitions be modified.
+Indeed, with appropriate logic in the bridge API, one palette may be able to support several distinct languages.
 
-Participants were very wary of including constant data into the source code directly.
-Many responses expressed general sentiment or specific preference for project-wide color theme or external resource file.
-Sometimes they explicitly suggested to add capability to handle the external resource files directly from the palette.
-Interestingly, most people seemed to consider regular expressions as part of the program logic.
-There were few people who complained about including the regular expression pattern string in the source code as opposed to the other two classes.
+\section{Lab Study}
+%\begin{figure*}[ht]
+%  \caption{test}
+%\end{figure*}
 
-%Examples of feedbacks in this category include:
+In order to evaluate the usefulness and usability of Graphite, we conducted a lab study with a palette designed for regular expressions (above).
+Participants were asked to write regular expressions, in the Java programming language, in response to several prompts.
+The control group was not given access to Graphite, but was otherwise free to use online resources.
+The treatment group was given a short tutorial of the Graphite system using a palette for the Color class as an example.
+The fact that a palette existed that was relevant to the tasks they would be asked to perform was mentioned, but the palette itself was not explicitly demonstrated, and its use was not required.
+
+Although our present sample size of 7 participants renders quantitative comparison difficult, several qualitative observations support the view that Graphite was helpful for users in this task.
+In particular, we found that the participants in the control group were facing difficulties that the palette was designed to address, including difficulty with the factory pattern used by Java regular expressions and the requirement that backslashes be doubly-escaped.
+We also observed that few subjects wrote adequate testing code to ensure that their regular expressions were correct.
+Participants in the treatment group, on the other hand, had few or no difficulties with these issues and wrote tests more frequently, due to the support for testing in the palette interface.
+Feedback from these participants, as well as participants in the control group who were shown the regular expression palette following the experiment, was also positive.
+
+%
+%\section{Future Work}
+%
+%We currently have a list of future works as following:
 %
 %\begin{itemize}
-%	\item The use of magic numbers in the source code is a bad design (54)
-%	\item Color picking is the `designer's job' (33)
-%	\item Would only use to define named constants (17)
-%	\item Would rather use color scheme / theme (3)
-%	\item Would use Object-Relational Mapping (ORM) / Language INtegrated Query (LINQ) instead of raw SQL query string (27 of 142)
-%	\item These palettes might be useful for prototyping, but not for real production (14)
-%	\item Instead of including the test cases into comments, they should be stored as unit tests (35 of 143)
+%	\item Implementing Graphite plug-in for different IDEs
+%	\item Improving the Graphite API for Javascript
+%	\item Implementing Keyword invocation model
+%	\item Evaluating Graphite tool more thoroughly.
 %\end{itemize}
-
-\subsubsection{Reversibility (19?)}
-
-Some participants expressed that it should be able to bring the palettes back when they want to modify the parameters after they used the palettes.
-
-\subsubsection{Palette Settings and State (41?)}
-
-Many participants wanted to be able to configure the palettes or have the palettes maintain state even after the palettes are hidden.
-% The followings are the comments that are related to this category.
-
-%\begin{itemize}
-%	\item Want recently used regular expressions, and control over the comments (12)
-%	\item Persistent database connection information is needed / want to manage connection pools (9)
-%	\item Want recent colors, capability of setting favorite colors (20)
-%\end{itemize}
-
-\subsubsection{Interaction with Code Context (13?)}
-
-Some people wanted to interact with the code context.
-For example, people expressed that it would be great if the SQL query palette was capable of dynamically creating SQL query string by combining multiple string variables, or by assigning variables for the parameters.
-
-\subsubsection{Performance (?)}
-
-Developers were concerned about the performance of these palettes.
-Some of the participants said that they would use these palettes only if it does not slow down the IDE.
-Also, this was the most popular comment on the reddit thread.
-
-\subsubsection{IDE Independence (?)}
-
-Several participants expressed a desire for IDE or even language independence for these palettes.
-
-\subsection{General Palette Design Considerations}
-
-General palette design considerations are about how to make more usable palettes.
-
-\subsubsection{Simplicity vs. Capability}
-
-Even though the participants were shown the same mockup screenshots, their reactions were split.
-Our color palette was considered too complex by many participants (26), and many others (26) would be satisfied with seeing colors next to a list of color names.
-The regular expressoin palette and SQL palette were considered too simple (12, and 15 participants respectively).
-They wanted syntax highlighting, match highlighting, and browsing capability of SQL databases.
-
-\subsubsection{Keyboard Navigability (12)}
-
-12 of the participants expressed strong antipathy toward the mouse. This was mostly directed at the color palette.
-
-\subsection{Suggestions from the Participants}
-At the end of the survey, we solicited suggestions from the participants about what types of classes could benefit from the active code completion palettes.
-119 participants suggested various types of classes and we classified the suggestions into several categories.
-
-\subsubsection{Alternative/Tricky/Literal Syntax (16)}
-These are the classes of which syntax is tricky, or complex.
-Even for a basic string class, people wanted to input multiline string or unicode string in more convenient way with automatic escaping.
-We address this issue in more detail in Section \#.
 %
-%\begin{itemize}
-%	\item Example classes
 %
-%	\begin{itemize}
-%		\item String
-%		\item Collections (e.g., dictionary)
-%		\item Vector, Matrix
-%		\item Embedded languages
-%		\item URLs, Paths
-%	\end{itemize}
-%	
-%	\item Expected features
-%	
-%	\begin{itemize}
-%		\item Proper escaping
-%		\item Syntax highlighting
-%		\item Auto-completion
-%	\end{itemize}
-%\end{itemize}
-
-\subsubsection{Unclear Parameter Implications (11)}
-The classes in this category has some parameters in it, and it is not easy to predict the results of the parameter modifications.
-Instead of running the whole application to see the result, the palette can serve as a preview panel so that the developers may check the results directly.
-Also, the palette can be a control panel which facilitates modifying the parameters as in Juxtapose [?].
-
-%\begin{itemize}
-%	\item Example classes
-%	
-%	\begin{itemize}
-%		\item Audio tweaks (modifying pitch, volume, etc.)
-%		\item 3D transformation matrices
-%		\item number / string / date formatting
-%		\item message box / input box (e.g., \texttt{JOptionPane})
-%	\end{itemize}
-%\end{itemize}
-
-\subsubsection{Query Languages (17)}
-Similar to the regular expression and SQL palette we presented, the developers also wanted to check if the other types of query strings (e.g., XPath / XQuery) are correct or not by checking the results directly on the palette.
-
-%\begin{itemize}
-%	\item Example classes
-%	
-%	\begin{itemize}
-%		\item Regular expression
-%		\item SQL query
-%		\item XPath / XQuery
-%	\end{itemize}
-%\end{itemize}
-
-\subsubsection{Graphical Elements (27)}
-The other popular category was graphical elements.
-We've already shown the color example, and the participants suggested many other graphical properties such as font, brush, shape and others.
-Also, people wanted to check or directly manipulate the GUI frames and layouts.
-
-%\begin{itemize}
-%	\item Example classes
-%	
-%	\begin{itemize}
-%		\item Color, Shape, Brush, Font, etc.
-%		\item JFrame, Layouts, any swing components
-%	\end{itemize}
-%\end{itemize}
-
-\subsection{Complex Instantiation Procedures (5)}
-Some of the participants pointed out that these palettes might be useful to create an object which requires complex instantiation code.
-For example, in order to read a text file in Java, the developer might want to use \texttt{BufferedReader} class.
-It is error prone to use this class because it requires try/catch block, and also one must close the file after reading it.
-By using a palette to choose a file or choose a variable which contains the file path, the developers could easily instantiate these objects.
-In the same way, it can alleviate the factory pattern usability problem \cite{ellis_factory_2007}.
-As long as the developers remember which class to use, they will not need to remember how to instantiate that class.
-
-%\begin{verbatim}
-%BufferedReader reader = null;
-%try {
-%    reader = new BufferedReader(
-%        new FileReader(filePath));
-%} catch (IOException e) {
-%    e.printStackTrace();
-%}
-%// use reader here
-%reader.close();
-%\end{verbatim}
-
-\subsection{Describe by Example (2)}
-Sometimes, it is possible to describe an object by examples.
-For instance, if there is a class which represents a shortcut key combination, we can easily instantiate an object of that class by pressing the actual shortcut key on the interactive palette.
-
-%\begin{itemize}
-%	\item Example classes
-%	
-%	\begin{itemize}
-%		\item Keyboard Keys
-%		\item Regular expression
-%	\end{itemize}
-%\end{itemize}
-
-\subsection{Integrating with Documentation, Tutorial (7)}
-Some participants suggested integrating the documentation or the tutorials into the interactive palettes.
-If the developers are not familiar of the specific API class, the palette could tell them how to use it.
-
-% conference papers do not normally have an appendix
-
-% use section* for acknowledgement
-%\section*{Acknowledgment}
-
-%The authors would like to thank...
 

### 2012-03-19 icse12/graphite-icse12.tex
Shortened some of the sentences.

@@
 \begin{center}
-\includegraphics[scale=0.8]{color_palette.png}\end{center}
+\includegraphics[width=\textwidth]{color_palette.png}\end{center}
 \caption{(a) An example code completion palette associated with the \texttt{Color} class. (b) The source code generated by this palette.}
@@
 The developer can interact with such palettes to provide parameters and other information related to her intent, and receive immediate feedback about the effect these choices will have on the behavior of the object being constructed.
-When this interaction is complete, the palette is responsible for generating appropriate source code for insertion at the cursor.
-Figure 1b shows the code that is inserted after the user presses the \textsc{enter} key.
+When this interaction is complete, the palette generates appropriate source code for insertion at the cursor.
+Figure 1b shows the inserted code after the user presses the \textsc{enter}.
 
@@
 
-Next, we describe Graphite, an Eclipse plug-in that provides active code completion for users of the Java programming language (Section 5).
-We describe several interesting design choices that were made to satisfy many of the requirements discovered in our preliminary investigations and briefly examine the trade-offs that were necessary due to Java's purely textual program representations and Eclipse's limited extension APIs.
+Next, we describe Graphite, an Eclipse plug-in that provides active code completion for the Java programming language (Section 5).
+We describe several design choices that were made to satisfy many of the requirements discovered in our preliminary investigations and briefly examine the trade-offs that were necessary due to Java's purely textual program representations and Eclipse's limited extension APIs.
 
@@
 A large number of participants volunteered comments: 193 for the color palette, 129 for the regular expression palette and 142 for the SQL palette.
-These responses were highly valuable when developing the design criteria below and helped to explain the patterns observed in Fig.
-\ref{frequencies}.
+These responses were highly valuable when developing the design criteria below and helped to explain the patterns observed in Figure \ref{frequencies}.
 
@@
 \subsection{Maintaining Separation of Concerns (183)}
-The most common issue developers had, particularly with the Color palette, was that it seemed to ignore the issue of separation of concerns by allowing developers to insert color constants directly into the program logic.
+The most common issue developers had, particularly with the Color palette, was that it ignored the issue of separation of concerns by allowing developers to insert color constants directly into the program logic.
 Many developers noted that this is considered bad practice, or should be limited to the prototyping phase of a project.
 This concern was also expressed in responses to the SQL palette, which required inserting parameters to connect to a database so that the query could be tested.
-The code that was inserted included initialization steps needed to connect to the specific database that was entered, and several participants noted that much of this information should appear in an external resource file rather than inline with the program logic itself.
+The resulting code included initialization steps needed to connect to the specific database that was entered, and several participants noted that much of this information should appear in an external resource file separated from the program logic.
 Few participants made similar comments about the regular expression palette, however, indicating that regular expressions are considered a part of the program logic rather than data by most developers.
@@
 \begin{itemize}
-	\item \verb|insert(str)|
-
-	Inserts the specified string at the cursor and closes the palette. The appropriate amount of indentation is automatically inserted after any newlines in this string.
-
-	\item \verb|cancel()|
-
-	Closes the palette without inserting a string.
-
-	\item \verb|getSelectedText()|
-
-	Returns the text that is currently selected by the user in the editor, or an empty string if no text has been selected. This method is used to implement reinvocation.
-
-	\item \verb|getIDE()|
-
-	Returns a string that specifies the IDE that is being used, ``Eclipse'' in our implementation.
-
-	\item \verb|getLanguage()|
-
-	Returns a string that specifies the programming language that is being used, ``java'' in our implementation.
+	\item \verb|insert(str)|: Inserts the specified string at the cursor and closes the palette. The indentation is automatically inserted after any newlines in this string.
+
+	\item \verb|cancel()|: Closes the palette without changes.
+
+	\item \verb|getSelectedText()|: Returns the text that is currently selected by the user in the editor, or an empty string if no text has been selected. This method is used to implement reinvocation.
+
+	\item \verb|getIDE()|: Returns a string that specifies the IDE that is being used, ``Eclipse'' in our implementation.
+
+	\item \verb|getLanguage()|: Returns a string that specifies the programming language that is being used, ``java'' in our implementation.
 \end{itemize}
@@
 Thus, he  spent several minutes looking up how to match whitespace correctly.
-In doing so, he found a description of the double-escape problem in Java and was able to fix the problem after that.
+In doing so, he found a description of the double-escape problem in Java and could fix the problem after that.
 \subsection{Treatment}
@@
 At other times, the subject used the external tool itself.
-The subject indicated that in some cases, he was not sure that our palette was free of bugs (when something unexpectedly matched or did not match\footnote{We did not observe any actual errors related to regex matching in our tool.}), and also because the external tool provided syntax and substring highlighting, a useful feature for complex regular expressions.
+The subject indicated that in some cases, he was not sure that our palette was free of bugs (when something unexpectedly matched or did not match\footnote{We observed no actual errors related to regex matching in our tool.}), and also because the external tool provided syntax and substring highlighting, a useful feature for complex regular expressions.
 When done with the external tool, the subject pasted the pattern into our tool to generate the appropriate code.
@@
 The other two subjects in the treatment group did not use any external tools.
-Both subjects decided not to use our palette for the first task (likely because they forgot that it was available, since they spent some time looking up API documentation after our initial demo with the color palette.) They both recognized the need for factory pattern instantiation and also both had an initial error related to escaping that they were able to resolve before moving on.
+Both subjects decided not to use our palette for the first task (likely because they forgot that it was available, since they spent some time looking up API documentation after our initial demo with the color palette.) They both recognized the need for factory pattern and also both had an initial error related to escaping that they were able to resolve before moving on.
 One of these subjects also began to write tests within a \verb|main| method.

### 2012-03-19 icse12/graphite-icse12.tex
Added some related work to icse12

--- 2012-03-18 icse12/graphite-icse12.tex
@@
 \usepackage{hyperref}
-\hypersetup{pdfborder={0 0 0 0}}
+\hypersetup{pdfborder={0 0 0}}
 
@@
 Several refinements and additions to the code completion menu have previously been suggested in the literature.
-These have focused on leveraging additional sources of information, such as databases of usage history \cite{robbes_how_2008}, examples extracted from code repositories \cite{bruch_learning_2009} and crowdsourced information \cite{mooty_calcite:_2010}, to increase the relevance and sophistication of the featured menu items.
+These have focused on leveraging additional sources of information, such as databases of usage history \cite{robbes_how_2008}\cite{HouPletcher2011}, inheritance information \cite{HouPletcher2011}, API-specific information \cite{HouPletcher2011}\cite{Lee+2008}, partial abbreviations \cite{Han+2009}, examples extracted from code repositories \cite{bruch_learning_2009} and crowdsourced information \cite{mooty_calcite:_2010}, to increase the relevance and sophistication of the featured menu items.
 As with the standard form of code completion, many of these sources of data can also be utilized via external tools (e.g.

### 2012-03-20 icse12/graphite-icse12.tex
Finished adding related work to icse12.

--- 2012-03-19 icse12/graphite-icse12.tex
@@
 Several refinements and additions to the code completion menu have previously been suggested in the literature.
-These have focused on leveraging additional sources of information, such as databases of usage history \cite{robbes_how_2008}, examples extracted from code repositories \cite{bruch_learning_2009} and crowdsourced information \cite{mooty_calcite:_2010}, to increase the relevance and sophistication of the featured menu items.
+These have focused on leveraging additional sources of information, such as databases of usage history \cite{robbes_how_2008}\cite{HouPletcher2011}, inheritance information \cite{HouPletcher2011}, API-specific information \cite{HouPletcher2011}\cite{Lee+2008}, partial abbreviations \cite{Han+2009}, examples extracted from code repositories \cite{bruch_learning_2009}\cite{Brandt+2010} and crowdsourced information \cite{mooty_calcite:_2010}\cite{SnipMatch}, to increase the relevance and sophistication of the featured menu items.
 As with the standard form of code completion, many of these sources of data can also be utilized via external tools (e.g.
@@
 \section{Related Work}
-In addition to the references provided in the introduction, some other concepts are related to active code completion.
-
 \subsection{Active Libraries}
@@
 
-Editing environments like Barista \cite{Ko06} has previously been developed that merge concepts from both conventional text based and structured editors.
-Barista allows developers to use alternative code representations.
-It differs from Graphite in that these representations are built directly into the editor (although adding new representations is possible), and that Barista requires the users to use its own model-view-controller based language, which is likely to be unfamiliar to many users.
+Editing environments like Barista \cite{Ko06} and the RBA editor \cite{RBA} also merge concepts from both text-based and structured editors by allowing for alternative code representations within a relatively conventional layout.
+Barista provides the opportunity for rich type-specific interfaces, but it is an IDE generation framework, so new extensions require recompilation.
+The RBA editor focuses on code readability rather than new modes of interaction, but new registrations can be added relatively easily.
+Both tools use a custom domain-specific language, which is likely to be unfamiliar to many users.
+
 In future work, we hope to explore an extensible, keyboard-driven, code-generation based approach that leverages structured code representations to eliminate the difficulties associated with reinvocation and maintaining palette state described in this paper.
@@
 There exist some IDE features that have been specifically designed for certain types.
-For example, CodeRush has a ``show color dialog'' \cite{CodeRush} which allows the developers to launch a color picker directly from the code editor.
-Resharper \cite{Resharper} also has a color assistance tool which previews the actual colors right beside the code completion menu items.
+For example, CodeRush \cite{CodeRush} and Resharper \cite{Resharper} have color dialogs that allow developers to launch a color picker directly from the code editor.
+IntelliJ IDEA has an inline regular expression palette, driven by its Intentions system, as well \cite{IntelliJRegexp}.
 However, these IDE specific features are hard-coded -- user-defined types cannot provide similar functionality.
-Recent versions of Visual Studio have started to explore user-defined palettes for widgets in the windowing system.
+Recent versions of Visual Studio support user-defined palettes associated with specific fields, rather than classes, of user interface widgets \cite{VSWidgets}.
+These are shown only in the property pane when using the graphical window layout editor.
 
@@
 \section*{Acknowledgment}
-
+% Funding sources
+% Class 
+% PLAID group / Jonathan Aldrich
+% Participants
+% Anonymous reviewers
+% UIUC people
+
+fasdfahsdfjk asdf alsdfhjkasdfh alsdhkjfh asdkflahs dfkjahs djkfhkjadfjakshf kjhasjdf akskdjfha sdhfkja sdfhajksdfa hsdfkaj sdfhasdf ahdfkajsd fhasdjkfhjas dkfa dfkadfa
+fasdfahsdfjk asdf alsdfhjkasdfh alsdhkjfh asdkflahs dfkjahs djkfhkjadfjak
 % trigger a \newpage just before the given reference

### 2012-03-20 icse12/graphite-icse12.tex
Cleaned up the references, added "palette reinvocation" in Section V, other small changes.

@@
 
+\usepackage{balance}
+
+\clubpenalty = 10000
+\widowpenalty = 10000
+\displaywidowpenalty = 10000
+
 \begin{document}
@@
 A fixed strategy determines the completions that are available, leaving library providers unable to specify new domain-specific or contextually-relevant items.
-In this paper we propose a technique called {\it active code completion} that eliminates these restrictions.
+In this paper we propose a technique called {\it active code completion} that eliminates these restrictions \cite{ACC_VLHCC}.
 This  makes developing and integrating a broad array of highly-specialized developer tools directly into the editor, via the familiar code completion command, significantly simpler.
@@
 \subsection{Survey Methods}
-We recruit participants for this survey\footnote{https://www.surveymonkey.com/s/2GLZP8V} primarily from a popular programming-related discussion forum hosted on the popular website reddit.com\footnote{http://www.reddit.com/r/programming}.
+We recruit participants for this survey\footnote{https://www.surveymonkey.com/s/2GLZP8V} primarily from a popular programming-related discussion forum hosted on the popular website reddit.com \cite{reddit_programming}.
 An additional 22 participants were recruited from the local computer science graduate student body.
@@
 
-We first asked participants about their level of familiarity with several programming languages, on a five-point Likert scale ranging from ``None'' to ``Expert''.
+We first asked participants about their level of familiarity with several programming languages, on a five-point Likert scale\footnote{``None'', ``Somewhat familiar'', ``Familiar'', ``Very familiar'', ``Expert''}.
 61.1\% of the participants indicated that they were an expert in at least one language, and an additional 35.7\% were ``very familiar'' with at least one language.
@@
 
-Although our presentation of active code completion did not include any completely mouse-driven interfaces, several participants commented that the Color palette included interface elements taken from standard color dialog boxes that could only be manipulated using the mouse.
-These comments were generally severe in their condemnation of mouse-based interfaces in developer tools, consistent with our finding that a significant portion of our participants were familiar with editors like Vim that place a strong emphasis on keyboard shortcuts.
+Although our mockup screenshots did not include any completely mouse-driven interfaces, several participants commented that the Color palette included interface elements taken from standard color dialog boxes that could only be manipulated using the mouse.
+These comments were generally severe in their condemnation of mouse-based interfaces in developer tools, consistent with our finding that a significant portion of our participants were using editors like Vim that place a strong emphasis on keyboard shortcuts.
 
@@
 
-The mockups we showed users were based on the Java language and the Eclipse development environment.
+The mockups we showed users were based on the Java and the Eclipse IDE.
 As we showed, a number of participants preferred other languages or editors.
-Many of these participants made comments asking that the features we describe not be restricted to one specific editor (e.g.
-NetBeans users) and some also called for support for other programming languages (e.g.
-Ruby and C\#).
+Many of these participants made comments asking that the features we describe be IDE and programming language independent.
 Indeed, the palettes we demonstrated could be used with only slight modifications in a variety of programming languages, given suitable architectural support for porting palettes between editing environments.
@@
 
-Our initial color palette was deemed overly complex by many participants (26), with many others (26) specifically saying that they would be satisfied with a simpler interface with colors listed next to their names.
+Our initial color palette was deemed overly complex by many participants (26), with many others (26) specifically wanting a simpler interface with colors listed next to their names.
 Other users wanted additional features, such as an ``eyedropper'' tool for selecting a color directly from an image on the screen.
@@
 Due to the wide range of user needs even within a single class, it may be that support for multiple palettes or a tabbed palette interface will be necessary in practice.
-To support incremental improvements to palettes based on user feedback, an architecture that makes deploying new and updated palettes relatively painless would also be valuable.
+To support incremental improvements based on user feedback, an architecture that makes deploying new and updated palettes relatively painless would also be valuable.
 
@@
 
+\subsection{Palette Reinvocation}
+To make further changes on the code generated by a Graphite palette, users can select the generated code and reinvoke the palette.
+The individual palette is responsible for parsing the selected code and filling the parameters back in the palette appropriately.
+When the user is satisfied with the new parameters and hits {\tt ENTER}, the selected code is replaced with a newly generated code.
+
 \subsection{Design Trade-Offs}
@@
 \item Because palettes are implemented in Javascript, any differences between Java and Javascript become evident.
-For example, Javascript color names are, in two cases, slightly different from the corresponding Java color name.
-The Javascript regular expression engine is also slightly different.
-Although a Java applet could be embedded to provide necessary interoperability in cases where these differences are critical, this is more difficult than it would be if the palettes were implemented using the same language as the code.
+For example, the color names are slightly different between Java and Javascript, and so are their regular expression engines.
+Although a Java applet could be used in cases where these differences are critical, this is still more difficult than it would be if the palettes were implemented using the same language as the code.
 \item The palette user interface must stay within its bounding box -- user interface elements like pop-up menus (unless they are provided by the browser itself) are thus more difficult to implement and may require additional API support in the future.
@@
 Implementing this modularly is particularly difficult, and the Eclipse API does not easily allow for  serialization of the sum of its contextual knowledge for consumption by a palette.
-\item Implementing a reinvocation mechanism that is robust to changes in the code and does not require the user to select the relevant information is difficult because this would require that the palette  parse all the code to find relevant information.
-This also would require richer insertion APIs that allow for  ``dead'' code to be removed.
+\item Implementing a robust reinvocation mechanism that does not require the user to select the generated code is difficult, because this would require the palette to parse all the code to find relevant information, and also remove the ``dead'' code.
 This puts a greater burden on palette developers and errors could affect user's code in subtle ways.
@@
 
-\subsection{Participants}
-
-We recruited 7 subjects from Computer Science Department and Institute for Software Research at Carnegie Mellon University.
-All the 7 subjects were PhD students.
-The subjects were randomly assigned into two groups: 4 subjects were assigned to the control group, and the other 3 subjects to the treatment group.
-There were six male participants and one female participant.
-Participants were compensated in the amount of 15 dollars for their participation.
-
 \subsection{Tasks}
@@
 The participants were allowed to move back and forth while doing the experiment, and they were allowed to use any external resources, including the internet, local console, and so forth (we did not observe any usage of programs other than the web browser, however).
-The only restriction we placed on their activity was that they were NOT allowed to directly search for the answer to a task online online.
+The only restriction we placed on their activity was that they were NOT allowed to directly search for the answer to a task online.
 We omit descriptions of each individual question due to space limitations.
@@
 %
-\subsubsection{Task Structure}
+%\subsubsection{Task Structure}
 
@@
 
-\section{User Study Results}
+\section{Pilot Study Results}
+
+\subsection{Participants}
+
+As a pilot study, we recruited 7 PhD students from School of Computer Science at Carnegie Mellon University.
+The subjects were randomly assigned into two groups: 4 subjects were assigned to the control group, and the other 3 subjects to the treatment group.
+There were six male participants and one female participant.
+Participants were compensated in the amount of 15 dollars for their participation.
+
+All participants had prior experience with both Java and regular expressions, but they were all somewhat rusty on the details of the API.
 After reading the prompt for the first task, all subjects began by searching for and reviewing documentation related to regular expressions.
@@
 This hybrid approach may help address some of the usability challenges previously associated with visual languages (cf.
-\cite{oai:CiteSeerXPSU:10.1.1.110.3146}).
+\cite{doi:10.1080/1049482940040202}).
 
@@
 %\bibliography{graphite-icse12}
-\bibliography{../research}
+\balance \bibliography{../research}
 

### 2012-03-20 icse12/graphite-icse12.tex
funding info. controlled user study -> pilot study

@@
 Using several empirical methods, we examine the contexts in which such a system could be useful, describe the design constraints governing the system architecture as well as particular code completion interfaces, and design one such system, named Graphite, for the Eclipse Java development environment.
-Using Graphite, we implement a palette for writing regular expressions as our primary  example and conduct a controlled user study that provides further evidence in support of the claim that integrating specialized tools directly into the editor environment is both feasible, particularly if aided by an  active code completion system like Graphite, and valuable to professional developers.
+Using Graphite, we implement a palette for writing regular expressions as our primary  example and conduct a small pilot study that provides further evidence in support of the claim that integrating specialized tools directly into the editor environment is both feasible, particularly if aided by an active code completion system like Graphite, and valuable to professional developers.
 \end{abstract}
@@
 
-Finally, we conduct a controlled lab study with a more complex palette, implemented using Graphite, that assists  developers as they write regular expressions (Section 6).
+Finally, we conducted a pilot lab study with a more complex palette, implemented using Graphite, that assists developers as they write regular expressions (Section 6).
 The data and observations gathered from this study provide specific evidence in support of the broader claim that highly-specialized tools that are integrated directly with the editing environment are particularly useful, and that an active code completion system like Graphite makes developing, deploying and discovering such tools fundamentally simpler.
 
-\section{Survey Results}
+\section{Survey}
 To validate our general conceptualization of active code completion, develop concrete design criteria and create a concrete list of use cases, we began by conducting a large survey of professional software developers.
@@
 %
-Next, we presented participants with a series of mockup palettes for a Color class (more complex than the one we ultimately implemented in Figure 1a), a regular expression class, and a SQL query class.
+We presented participants with a series of mockup palettes for a Color class (more complex than the one we ultimately implemented in Figure 1a), a regular expression class, and a SQL query class.
 Participants were also shown mockup screenshots demonstrating how a user could invoke the palette, and a mockup showing the code that would be inserted once a selection had been made.
@@
 
-fasdfahsdfjk asdf alsdfhjkasdfh alsdhkjfh asdkflahs dfkjahs djkfhkjadfjakshf kjhasjdf akskdjfha sdhfkja sdfhajksdfa hsdfkaj sdfhasdf ahdfkajsd fhasdjkfhjas dkfa dfkadfa
-fasdfahsdfjk asdf alsdfhjkasdfh alsdhkjfh asdkflahs dfkjahs djkfhkjadfjak
+Funding for this work comes in part from the Korea Foundation for Advanced Studies (KFAS) and in part from NSF grant CCF-0811610.
+Any opinions, findings and conclusions or recommendations expressed in this material are those of the author(s) and do not necessarily reflect those of KFAS or the NSF.
+
 % trigger a \newpage just before the given reference

### 2012-03-21 icse12/graphite-icse12.tex
Final ICSE2012 paper.

@@
 Refinements to the code completion menu that incorporate additional sources of information have similarly been shown to be valuable, even relative to standalone counterparts offering similar functionality.
-In this paper, we describe {\em active code completion}, a technique that enables the integration of a much broader array of interactive and highly-specialized code generation interfaces, called {\em palettes}, directly into the editor environment.
+In this paper, we describe {\em active code completion}, an architecture that allows library developers to introduce interactive and highly-specialized code generation interfaces, called {\em palettes}, directly into the editor.
 Using several empirical methods, we examine the contexts in which such a system could be useful, describe the design constraints governing the system architecture as well as particular code completion interfaces, and design one such system, named Graphite, for the Eclipse Java development environment.
-Using Graphite, we implement a palette for writing regular expressions as our primary  example and conduct a small pilot study that provides further evidence in support of the claim that integrating specialized tools directly into the editor environment is both feasible, particularly if aided by an active code completion system like Graphite, and valuable to professional developers.
+Using Graphite, we implement a palette for writing regular expressions as our primary  example and conduct a small pilot study.
+In addition to showing the feasibility of this approach, it provides further evidence in support of the claim that integrating specialized code completion interfaces directly into the editor is valuable to professional developers.
 \end{abstract}
@@
 Calcite \cite{mooty_calcite:_2010} uses information that could already be accessed using the Jadeite \cite{conf/vl/StylosFYM09} tool).
-Empirical evidence presented in these studies, however, indicates that directly integrating these kinds of tools directly into the editor environment can be more effective.
+Empirical evidence presented in these studies, however, suggests that directly integrating these kinds of tools into the editor is particularly effective.
+For example, users of the Calcite tool completed 40\% more tasks in a lab study (unfortunately, a Jadeite control group was not included.)
 
@@
 When an item is selected, code is inserted immediately, without further input from the developer.
-These systems are also limited in that they are difficult to modularly extend.
-A fixed strategy determines the completions that are available, leaving library providers unable to specify new domain-specific or contextually-relevant items.
-In this paper we propose a technique called {\it active code completion} that eliminates these restrictions \cite{ACC_VLHCC}.
+These systems are also difficult to extend: a fixed strategy determines the completions that are available, so library providers cannot directly specify new domain-specific or contextually-relevant logic.
+In this paper we propose a technique called {\it active code completion} that eliminates these restrictions\footnote{Portions of this work previously appeared in a poster abstract \cite{ACC_VLHCC}.}.
 This  makes developing and integrating a broad array of highly-specialized developer tools directly into the editor, via the familiar code completion command, significantly simpler.
@@
 
-If the developer invokes the code completion command at the indicated cursor position, the editor  looks  for a {\it palette definition} associated with the {\it type} of the expression being entered, which in this case is  \verb|Color|.
+If the developer invokes the code completion command at the indicated cursor position (\textvisiblespace), the editor  looks  for a {\it palette definition} associated with the {\it type} of the expression being entered, which in this case is  \verb|Color|.
 If an associated palette is found, a menu item briefly describing this palette is added to the standard code completion menu.
 When selected, the corresponding palette is shown, replacing the standard code completion menu.
-Figure 1a gives an example of a simple palette that may be associated with the \verb|Color| class.
+Figure 1a gives an example of a simple palette that may be associated with the \verb|Color| class\footnote{A video demonstrating this process is available from the webpage above.}.
 
@@
 When this interaction is complete, the palette generates appropriate source code for insertion at the cursor.
-Figure 1b shows the inserted code after the user presses the \textsc{enter}.
-
-In accordance with rational design practices, we sought to address the following questions before designing and implementing our active code completion system:
+Figure 1b shows the inserted code after the user presses \textsc{enter}.
+
+In accordance with best practices, we sought to address the following questions before designing and implementing our active code completion system:
 
@@
 
-To help us answer these questions, we conducted a large online survey of professional developers (Section 2).
-Their responses, along with information gathered from informal interviews and code corpus analyses, revealed a number of interesting use cases (Section 3) and non-trivial functional requirements for palette interfaces as well as the underlying active code completion architecture (Section 4).
-Notably, many of these findings can be applied to development tools for professional developers more generally, not simply those integrated into the editor in the way we have shown.
-
-Next, we describe Graphite, an Eclipse plug-in that provides active code completion for the Java programming language (Section 5).
-We describe several design choices that were made to satisfy many of the requirements discovered in our preliminary investigations and briefly examine the trade-offs that were necessary due to Java's purely textual program representations and Eclipse's limited extension APIs.
-
-Finally, we conducted a pilot lab study with a more complex palette, implemented using Graphite, that assists developers as they write regular expressions (Section 6).
-The data and observations gathered from this study provide specific evidence in support of the broader claim that highly-specialized tools that are integrated directly with the editing environment are particularly useful, and that an active code completion system like Graphite makes developing, deploying and discovering such tools fundamentally simpler.
+To help us answer these questions, we conducted a survey of 473 professional developers (Section II).
+Their responses, along with information gathered from informal interviews and code corpus analyses, revealed a number of non-trivial functional requirements for palette interfaces as well as the underlying active code completion architecture (Section III).
+Participants also suggested a large number of use cases, demonstrating the broad applicability of this technique.
+We organize these into several broad categories (Section IV).
+
+Next, we describe Graphite, an Eclipse plug-in that implements the active code completion architecture for the Java programming language (Section V), allowing Java library developers to associate custom palettes with their own classes.
+We describe several design choices that we made to satisfy the requirements discovered in our preliminary investigations and briefly examine necessary trade-offs.
+
+Finally, we conducted a pilot lab study with a more complex palette, implemented using Graphite, that assists developers as they write regular expressions (Section VI).
+The study provides specific evidence in support of the broader claim that highly-specialized tools that are integrated directly with the editing environment are particularly useful.
+We conclude that active code completion systems like Graphite are useful because they make developing, deploying and discovering such tools fundamentally simpler.
 
 \section{Survey}
-To validate our general conceptualization of active code completion, develop concrete design criteria and create a concrete list of use cases, we began by conducting a large survey of professional software developers.
+To validate our general conceptualization of active code completion, develop concrete criteria to constrain our system and palette designs, and create a list of use cases to justify this effort, we began by conducting a large survey of professional software developers.
 
@@
 By doing this, the respondents could understand the concept precisely and give useful feedback relevant to the above questions.
-\subsection{Survey Methods}
-We recruit participants for this survey\footnote{https://www.surveymonkey.com/s/2GLZP8V} primarily from a popular programming-related discussion forum hosted on the popular website reddit.com \cite{reddit_programming}.
-An additional 22 participants were recruited from the local computer science graduate student body.
+\subsection{Participants}
+We recruited participants for this survey\footnote{https://www.surveymonkey.com/s/2GLZP8V} primarily from a popular programming-related discussion forum hosted on the popular website reddit.com \cite{reddit_programming}.
+An additional 22 participants were computer science graduate students at CMU.
 
@@
 Few participants (4\%) indicated that they were unfamiliar with regular expressions and no participants were unfamiliar with creating SQL queries, providing further evidence that our participants were not novice developers.
-Fig.
-\ref{strategies} summarizes the strategies that participants generally preferred for instantiating regular expressions and SQL queries.
+Figure \ref{strategies} summarizes the strategies that participants generally preferred for instantiating regular expressions and SQL queries.
 Using an external tool was a common strategy in both cases, particularly for SQL queries, but several other strategies were also represented.
@@
 Finally, after showing the series of mockup screenshots, we asked participants to rate how useful the integrated palette would be to them if they needed to instantiate the corresponding class.
-The responses to this question for each palette are summarized in Fig.
-\ref{frequencies}.
+The responses to this question for each palette are summarized in Figure \ref{frequencies}.
 In each case, more than half of the participants indicated that they would use active code completion at least some of the time.
@@
 \section{Design Criteria}
-Using the information gathered from the survey as well as informal discussions with developers and researchers over the course of this study, we sought to develop design criteria constraining both the overall system design as well as the design of individual palettes.
+Using the information gathered from the survey as well as informal discussions with developers and researchers, we developed  design criteria constraining both the overall system design as well as the design of individual palettes.
 In the section headings below, the number of survey responses, summed over the three palette mockups, that contained the listed concern, as judged by the authors of this paper, are listed in parenthesis.
-Many of these concerns apply generally to various kinds of editor-integrated tools, not just those implemented using the active code completion mechanism that we focus on, so we hope these design criteria will be valuable to the developer tool design community more broadly.
+These criteria were useful in designing Graphite (Section V) and we note that this collection of criteria may also be relevant to researchers designing other kinds of editor-integrated tools.
+Based on the variety of concerns expressed by participants in our survey, the design space for these tools appears to be quite complex.
 
 \subsection{Maintaining Separation of Concerns (183)}
-The most common issue developers had, particularly with the Color palette, was that it ignored the issue of separation of concerns by allowing developers to insert color constants directly into the program logic.
+The most common issue participants had was that palettes seemed to violate the principle of separation of concerns.
+The color palette, for example, allows developers to insert color constants directly into the program logic.
 Many developers noted that this is considered bad practice, or should be limited to the prototyping phase of a project.
@@
 
-This feedback suggests that tool and palette designers may wish to explicitly aprise users of relevant best practices and acknowledge that palettes that generate constant data may be most useful in the prototyping phase.
-It can be noted that when transitioning from a prototype to production-quality software the code generated by a palette or tool may be used as template that can be refactored to follow stricter policies as needed.
-We showed users the Color and SQL palettes in the context of a standard programming editor, but these comments indicate that resource file and  stylesheet editors may benefit from active code completion support as well.
+This suggests that tool and palette designers may wish to explicitly aprise users of relevant best practices and acknowledge that palettes that generate constant data may be most useful in the prototyping phase.
+It can be noted that when transitioning from a prototype to production-quality software the code generated by a palette or tool may be used as template to be refactored as needed.
+It also suggests that resource file and stylesheet editors may particularly benefit from active code completion support.
 
@@
 These test strings and the results of performing the match were inserted as comments below the generated source code.
-A number of participants requested that explicit unit tests be generated instead.
-Although the palette interface conveyed the results of these tests immediately, participants remained adamant about explicit unit tests, likely due to concerns that future modifications to the generated code without reinvoking the palette might introduce bugs, or due to the desire to conform to standard testing practices.
-To support the generation of unit tests, however, the active code completion architecture must explicitly allow for code generation at locations other than at the cursor and handle the fact that statements generated after a code completion command are not necessarily encapsulated well enough to be immediately testable in isolation.
+A number of participants requested that unit tests be generated instead, likely due to concerns that future modifications might introduce bugs, or due to the desire to conform to standard testing practices.
+To support the generation of unit tests, the active code completion architecture would need to support code generation at locations other than at the cursor.
 
@@
 
-A related feature important to many participants is support for maintaining settings and usage history across invocations of the same palette at different code locations.
-For example, 20 participants requested that the Color palette include a list of recent or favorite colors, and 12 participants specifically inquired about whether the database connection information was maintained between invocations of the SQL palette.
+A related feature important to many participants was support for maintaining settings and usage history across invocations of the same palette at different code locations.
+For example, 20 participants requested that the Color palette include a list of recent or favorite colors, and 12 participants inquired about whether the database connection information was maintained between invocations of the SQL palette.
 
@@
 
-\subsection{Minimal Impact on Editor Responsiveness}
+\subsection{Responsiveness}
 
@@
 
-Our initial color palette was deemed overly complex by many participants (26), with many others (26) specifically wanting a simpler interface with colors listed next to their names.
-Other users wanted additional features, such as an ``eyedropper'' tool for selecting a color directly from an image on the screen.
-Similarly, the regular expression and SQL palettes were considered too simple by 12, and 15 participants respectively.
-User requests included syntax highlighting, a range of additional helpful tools for generating, testing and sharing regular expressions, and the ability to browse databases more directly.
-Due to the wide range of user needs even within a single class, it may be that support for multiple palettes or a tabbed palette interface will be necessary in practice.
-To support incremental improvements based on user feedback, an architecture that makes deploying new and updated palettes relatively painless would also be valuable.
+While some participants wanted simpler palettes, others requested significant new capabilities, indicating that user needs may vary substantially.
+For example, our initial color palette (different from the one shown in Figure 1) was deemed overly complex by many participants (26).
+Other users wanted additional features, such as an ``eyedropper'' tool (11) for selecting a color directly from an image on the screen.
+The regular expression and SQL palettes were considered too simple (by 12, and 15 participants respectively).
+Users suggested syntax highlighting, several mechanisms for generating, testing and sharing regular expressions and queries, and the ability to browse SQL databases.
+
+Due to this wide range of user needs even for a single class, it may be that support for multiple palettes or a tabbed palette interface would be helpful in practice.
+To support incremental improvements based on such user feedback, an architecture that makes deploying new and updated palettes relatively painless would also be valuable.
 
 \section{Use Cases}
-At the end of the survey, we solicited suggestions from the participants of other classes that could benefit from an associated palette.
-119 participants made one or more suggestion and we classified the suggestions into several categories.
-We also mention other suggestions made by researchers and developers via other channels in some cases, without including them in the provided counts.
+At the end of the survey, we asked the the participants to suggest other classes that could benefit from an associated palette to support our claim that active code completion is broadly applicable, and to allow us to characterize the specific scenarios where it may be most useful.
+A total of 119 participants made one or more suggestions, which we classified into several broad categories (we omit a few of these below due to space constraints).
+As above, the number of participants suggesting a palette in each category is listed in parenthesis.
+We also include suggestions made by researchers and developers in private discussions without including them in the provided counts.
 
@@
 
-\subsection{Proof Assistants (1)}
+\subsection{Proof Assistants}
 A proof assistant is a tool for constructing proof terms.
@@
 Active code completion works directly with types to help developers construct expressions, so if applied to a language with a cleanly-developed connection to formal logic (e.g.
-Coq), palettes would be useful for constructing highly interactive proof assistant interfaces.
+Coq), palettes would be useful for constructing  interactive proof assistant interfaces.
 
@@
 %
-\section{Tool Design}
-Motivated by these use cases and with the design criteria developed as a result of our developer survey in mind, we sought to create an active code completion system.
-We named our project Graphite, an acronym for {\bf Gra}phical {\bf P}alettes to {\bf H}elp {\bf I}nstantiate {\bf T}ypes in the {\bf E}ditor.
-We describe the design decisions and trade-offs that went into this system in this section.
-A video demonstrating Graphite, including the process of writing a full palette, is available for further details \footnote{http://www.youtube.com/watch?v=FfCoXQEHGAI}.
-The full source code of Graphite is also available at the URL given on the first page.
-
-We chose to use Eclipse for the Java programming language because this combination was the most widely-used  amongst participants in our survey.
-Eclipse plug-in development is well-supported and there are many examples available online that we can refer to, and the Eclipse project is available under an open source license.
-
-\subsection{HTML+Javascript based palettes}
-Eclipse is itself written in Java and uses the SWT and JFace graphical user interface toolkits.
-These frameworks are not widely used outside of the Eclipse ecosystem.
-As we have noted, a number of participants in our field study indicated that they hoped that our tool would be available for other IDEs and other programming languages.
-Indeed, many concepts (including regular expressions, colors and database queries) are similar across languages.
-To better support IDE-independence and language-independence, we chose instead to build palettes using HTML and Javascript.
-Javascript was among the most well-known languages in our study, just behind Java, C and C++, and is highly flexible.
-A number of user interface libraries are available, such as jQuery\footnote{http://jquery.com/} (which we used to implement our palettes).
-All major browsers now feature sophisticated debuggers and run-time inspection facilities that ease debugging and testing.
-Deploying palettes using standard URLs is also simpler than attempting to integrate their source code directly into Java libraries and packages and eases the process of incremental development and deployment.
-All major windowing toolkits feature a web browser control that can allow for the display of these palettes within the editor environment.
-In Eclipse, we used the standard SWT Browser control.
+\section{System Design and Implementation}
+After completing the survey, we built an active code completion system named Graphite, an acronym for {\bf Gra}phical {\bf P}alettes to {\bf H}elp {\bf I}nstantiate {\bf T}ypes in the {\bf E}ditor. 
+We chose to build the system as an extension to Eclipse for Java because this combination was the most widely-used  amongst participants in our survey.
+In the subsections below, we describe how several novel design decisions made it possible to satisfy design criteria from Section III and enabled several use cases described in Section IV.
+The end result is a simple system that allows an API's developers, as well as external developers, to build rich HTML5-based palettes that can be associated with both in-built and user-defined classes directly.
+Developers using an API can discover and invoke these palettes through the standard code completion menu.
+
+\subsection{HTML5-Based Palettes}
+Palette developers build palettes using HTML5 technologies (HTML, CSS and JavaScript). We made this decision for several reasons. 
+Eclipse is written in Java and uses the SWT and JFace graphical user interface toolkits, but these are not widely used outside of the Eclipse ecosystem.
+JavaScript was among the most well-known languages in our study, just behind Java, C and C++, and is highly flexible.
+A number of useful libraries are available (e.g.
+jQuery \cite{jQuery}.)
+
+As noted in Section IV-H, a number of participants in our field study indicated that they hoped that our tool would be available for other IDEs and other programming languages.
+All major windowing toolkits feature a web browser control, so HTML5-based palettes can easily be loaded by different editor environments without developer intervention.
+
+Deploying palettes using standard URLs is also simpler than attempting to integrate them directly into Java libraries and packages.
+It also eases the process of incremental and rapid development, as all major browsers now feature sophisticated debuggers and run-time inspection facilities.
 
 \subsection{Palette API}
-In order to allow palettes to communicate with the surrounding editor environment, a simple Javascript interface was injected.
-These functions were implemented by the editor plug-in, but appear as ``native'' browser functions to Javascript code.
+Palettes communicate with the host IDE using a simple Javascript API.
+To access this API, the palette must include a small script named \verb|graphite.js| into their page.
+The default implementation of this script provides an implementation of the API methods for testing in a standard web browser.
+When loaded into an editor, the editor plug-in replaces these methods with specialized implementations.
 The API consists of the following methods, accessed through a global object named \verb|graphite|:
@@
 
-	\item \verb|cancel()|: Closes the palette without changes.
-
-	\item \verb|getSelectedText()|: Returns the text that is currently selected by the user in the editor, or an empty string if no text has been selected. This method is used to implement reinvocation.
-
-	\item \verb|getIDE()|: Returns a string that specifies the IDE that is being used, ``Eclipse'' in our implementation.
-
-	\item \verb|getLanguage()|: Returns a string that specifies the programming language that is being used, ``java'' in our implementation.
+	\item \verb|cancel()|: Closes the palette without inserting code. Note that the ESC key cancels by default, and palettes are modal so clicking outside a palette will not cancel.
+
+	\item \verb|getSelectedText()|: Returns the text that is currently selected in the editor, or an empty string if no text has been selected.
+This method is used to implement reinvocation -- users simply select previously generated code and invoke the palette as described in Section I.
+The palette is responsible for parsing the selected text to extract relevant parameters.
+
+	\item \verb|getIDE()|, \verb|getLanguage()|: Return a string that specifies the IDE and language that is being used, ``Eclipse'' and ``Java'' in our implementation.
 \end{itemize}
@@
 By limiting the complexity of this API, we reasoned that developers would be able to create specialized palettes more easily.
-An additional benefit is that the developers of other editing environments are also able to create plug-ins that support Graphite palettes with minimal effort.
-
-To access this API, the palette must include a small script named \verb|graphite.js| into their page.
-This file interacts with the host editor to create the graphite object and provides an alternative implementation of these methods when the palette is being tested in a web browser.
+A simple hello world palette is only two lines of code, for example.
+An additional benefit is that the developers of other editing environments should also be able to create plug-ins that support Graphite palettes with minimal effort.
 
@@
 \begin{figure*}\begin{center}
+\vspace{2mm}
 \includegraphics[scale=.6]{regex.png}\end{center}
@@
 
-Graphite currently provides two methods for associating a palette with a particular type so that the Graphite plug-in is able to discover it when the code completion menu is invoked by a developer.
+Graphite currently provides two methods for associating a palette with a particular class so that the editor plug-in can include it in the code completion menu when relevant.
 
 \subsubsection{Annotation-based}
-For user-defined classes that the palette author has the authority to modify, the \texttt{@GraphitePalette} annotation associates a palette with the class.
+For user-defined classes that the palette developer has the authority to modify, the \texttt{@GraphitePalette} annotation associates a palette with the class.
 The annotation must specify the URL of the palette, and can contain some other optional information (e.g.
 the description that is shown in the code completion menu).
-This  allows library providers to provide palettes that are specialized to their libraries and distribute them alongside their code.
-In this way, the users of the libraries are not required to discover that an external tool exists.
-If the user does not have the Graphite plug-in for Eclipse installed, the annotation has no effect.
+This  allows API providers to provide palettes that are specialized to their libraries and distribute them directly alongside their code.
+The benefit of this approach is that users of the API are not required to discover that an external tool exists and explicitly install it into their IDE.
 
 \subsubsection{Explicit}
-In cases where the palette developer cannot modify a class directly (such as palettes for classes in the Java standard library), a second method exists to explicitly associate a palette with the fully qualified name of a class.
-Currently, this feature is exposed as a preference pane in the Eclipse IDE.
-To facilitate automated installs in the future, a preferences file may also be added to the system.
-
-\subsection{Palette Reinvocation}
-To make further changes on the code generated by a Graphite palette, users can select the generated code and reinvoke the palette.
-The individual palette is responsible for parsing the selected code and filling the parameters back in the palette appropriately.
-When the user is satisfied with the new parameters and hits {\tt ENTER}, the selected code is replaced with a newly generated code.
+In cases where a palette developer cannot modify a class directly (such as palettes for classes in the Java standard library), end-users can explicitly associate a palette with the fully qualified name of a class via a preference pane in the Eclipse IDE.
 
 \subsection{Design Trade-Offs}
-The design that we have described is light-weight, highly flexible and highly responsive.
-It lays foundations for IDE and language portability.
-However, several trade-offs have also been made to achieve these benefits.
+The design that we have described is light-weight, highly flexible and does not significantly impact IDE responsiveness.
+It also lays foundations for IDE and language portability.
+However, this design also leads to trade-offs:
 
 \begin{itemize}
-\item Because palettes are implemented in Javascript, any differences between Java and Javascript become evident.
-For example, the color names are slightly different between Java and Javascript, and so are their regular expression engines.
-Although a Java applet could be used in cases where these differences are critical, this is still more difficult than it would be if the palettes were implemented using the same language as the code.
+\item Because palettes are implemented in Javascript, any differences between the semantics of Java and Javascript can be problematic.
+For example, color names are slightly different between Java and Javascript, as are the regular expression engines.
+Although a Java applet could be used in cases where these differences are critical, this is still more difficult than it would be if the palettes were implemented in the same language.
 \item The palette user interface must stay within its bounding box -- user interface elements like pop-up menus (unless they are provided by the browser itself) are thus more difficult to implement and may require additional API support in the future.
@@
 Implementing this modularly is particularly difficult, and the Eclipse API does not easily allow for  serialization of the sum of its contextual knowledge for consumption by a palette.
-\item Implementing a robust reinvocation mechanism that does not require the user to select the generated code is difficult, because this would require the palette to parse all the code to find relevant information, and also remove the ``dead'' code.
-This puts a greater burden on palette developers and errors could affect user's code in subtle ways.
-The small inconvenience of having to select code before reinvoking a palette may be balanced by the knowledge that no other portions of the code will be modified inadvertently by that palette.
+\item Because the reinvocation mechanism relies on parsing the selected code, the burden is high for both palette developers (although Javascript libraries for parsing Java code are available) and palette users.
+A better solution would be one where palette-related metadata is stored directly with the generated code.
+This could be partially addressed by the use of special comments to delimit sections of generated code and store palette metadata, but a more elegant solution would require a  change in how Java source code is represented.
 \end{itemize}
 
-\subsection{Palettes}
-We implemented two full-featured palettes to demonstrate and test the functionality of this framework: a regular expression palette (Figure 5) and a color selection palette (Figure 1).
-
-\subsubsection{Regular Expressions}
-Writing correct regular expression patterns is difficult.
-The regular expression palette, associated with the \verb|Pattern| class in the Java standard library, allows users to enter and test regular expression patterns interactively before inserting them into their code.
-The focal point of this palette is the pattern input area.
-As the user enters a pattern into this input area, syntax errors are indicated with a red background; the background remains white when the pattern is valid.
-
-In addition to the pattern, a regular expression consists of flags that can change its matching behavior in various ways.
-Our palette allows users to toggle the case-sensitivity flag of the regular expression using a small checkbox labeled \verb|Ignore Case|, placed next to the input area.
-A keyboard shortcut is also available, indicated using the underlined letter (Ctrl+I).
-
-To allow users to test the behavior of the regular expression that they have entered, two columns are available below the expression input area.
-The left column contains an input area labeled ``Should match...'' and the right column contains an input area labeled ``Should NOT match...''.
-Users use these to enter lists of test  strings into each column.
-The background colors behind these strings change to indicate whether the regular expression that has been entered matches or does not match that string.
-Green indicates that the pattern  {\it matched} the string and red indicates that it does {\it not} match, regardless of the column that the test is in (this scheme was chosen based on feedback from an initial pilot study of this palette.) A key describing this color scheme is displayed after the first test has been entered (not shown).
-
-Users navigate between the three input areas using standard \verb|Tab| cycling behavior.
-The label in each text area remains visible until some input has been entered, rather than disappearing immediately on focus.
-When a user is satisfied with the regular expression that they have entered, she can press the \verb|Enter| key to insert the appropriate Java source code.
-Because Java requires that the regular expression pattern be placed inside a string literal, additional escape sequences are needed in front of backslashes.
-This can be tedious and error-prone if done manually.
-Because the palette is responsible for generating this literal, however, this is no longer a problem.
-In addition to the source code itself, the tests are retained in a comment beginning on the next line.
-The user can leave these as documentation or use them to create more formal unit tests within their preferred testing framework.
-
-If the user wishes to modify the regular expression or change the test set, she can highlight the code that was inserted and then invoke the palette once again.
-The palette parses the selected text to extract the regular expression and tests.
-Although this is slightly inconvenient in that it requires that the user select the relevant portion of the code, it is the most robust method for implementing this feature -- it can easily handle changes to the regular expression or tests and it does not require additional logic in the palette to determine which portions of the code are relevant, and should subsequently be replaced if a new regular expression is entered.
-
-\subsubsection{Color Selection}
-We omit a description of the full behavior of the Color selection palette due to space limitations. 
 %\section{Future work}
@@
 
-\section{User Study Design}
-
-We conducted a comparative study to evaluate the usefulness and usability of a Graphite palette in a constrained setting -- writing regular expressions.
-We used a between-subjects design by randomly splitting the participants into two groups: a control group, where subjects were not shown nor able to use the palette, and a treatment group, where subjects were shown a different palette and allowed to discover and use the regular expression palette if they wished.
-
-\subsection{Process}
-\subsubsection{Preliminary Survey}
-The participants were first asked to fill out the pre-survey form about their prior experiences with various programming languages, integrated development environments (IDEs), and regular expressions, based on the previous survey that we had conducted.
-Of note, the subjects in this study were considerably less skilled with Java and regular expressions than in the previous survey, likely due to a sampling effect -- our subjects were all PhD students at Carnegie Mellon.
+\subsection{Palettes}
+We implemented two palettes using the jQuery library for basic functionality \cite{jQuery}, which we used in the pilot study described below.
+
+\subsubsection{Color Selection}
+The color selection palette we ultimately developed was significantly simpler than the one shown in the preliminary survey, due to user comments.
+It allows users to enter any valid CSS color string, satisfying the requirements for keyboard navigability.
+The entered color is syntax checked and a preview is shown.
+Standard Java colors are also available as swatches that can be selected by the mouse if desired.
+Reinvocation support is provided.
+This palette took about 500 lines of code and markup, not including the jQuery library.
+Much of this was needed to translate arbitrary CSS color strings into RGB values, rather than for user interface logic.
+
+\subsubsection{Regular Expressions}
+Writing correct regular expression patterns is difficult.
+The regular expression palette, associated with the \verb|Pattern| class in the Java standard library, allows users to enter and test regular expression patterns interactively before inserting them into their code.
+The focal point of this palette is the pattern input area.
+As the user enters a pattern into this input area, syntax errors are indicated with a red background; the background remains white when the pattern is valid.
+
+In addition to the pattern, a regular expression consists of flags that can change its matching behavior in various ways.
+Our palette allows users to toggle the case-sensitivity flag of the regular expression using a small checkbox labeled \verb|Ignore Case|, placed next to the input area.
+A keyboard shortcut is also available, indicated using the underlined letter (Ctrl+I).
+
+To allow users to test the behavior of the regular expression that they have entered, two columns are available below the expression input area.
+The left column contains an input area labeled ``Should match...'' and the right column contains an input area labeled ``Should NOT match...''.
+Users use these to enter lists of test  strings into each column.
+The background colors behind these strings change to indicate whether the regular expression that has been entered matches or does not match that string.
+Green indicates that the pattern  {\it matched} the string and red indicates that it does {\it not} match, regardless of the column that the test is in (this scheme was chosen based on feedback from an initial pilot of this palette, the results of which are not included below.) A key describing this color scheme is displayed after the first test has been entered (not shown above, see video).
+
+Users can navigate the palette using the keyboard using standard \verb|Tab| cycling behavior.
+The label in each text area remains visible until some input has been entered, rather than disappearing immediately on focus.
+When a user is satisfied with the regular expression that they have entered, she can press the \verb|Enter| key to insert the appropriate Java source code.
+Because Java requires that the regular expression pattern be placed inside a string literal, additional escape sequences are needed in front of backslashes.
+This can be tedious and error-prone if done manually.
+The palette automatically inserts these escape sequences.
+In addition to the source code itself, the tests are retained in a comment beginning on the next line.
+If the user wishes to modify the regular expression or change the test set, she can highlight the code that was inserted and then invoke the palette once again.
+The palette parses the selected text to extract the regular expression and tests.
+
+This palette required about 700 lines of code and markup.
+
+\section{Pilot Study}
+
+We conducted a small controlled pilot study to evaluate the usefulness and usability of the Graphite system for a specific development task -- writing regular expressions -- and found significant benefits for the treatment group.
+
+\subsection{Study Methods}
+\subsubsection{Between-Subjects Design}
+We used a between-subjects design by randomly assigning the participants to either the control group or the treatment group.
+In the control group, subjects were not shown or able to use any palettes.
+In the treatment group, subjects were shown the simple color palette shown in Figure 1, and allowed to discover and use the regular expression palette shown in Figure 5 if they wished.
+No specific training on the use of this palette was provided, to simulate realistic usage scenarios.
+
+We chose a \textit{between-subjects design} because a within-subjects design would have required us to produce pairs of tasks with equal difficulties.
+This turned out to be very challenging.
+Second, we could not easily ignore a learning effect during the experiment if we had used a within-subjects design.
+We observed that this effect was quite strong, as most subjects had not used regular expressions recently.
 
@@
 
-\subsection{Within vs. between-subjects design}
-
-We decided to use a \textit{between-subjects design} because a within-subjects design would have required us to produce pairs of tasks with equal difficulties.
-This turned out to be very difficult.
-Second, we could not easily ignore a learning effect during the experiment if we had used a within-subjects design.
-We observed that this effect was quite strong, as most subjects hadn't used regular expressions recently.
-
-\subsection{Tasks}
+\subsubsection{Tasks}
 
@@
 
+\subsection{Participants}
+
+We recruited 7 PhD students from CMU.
+The subjects were randomly assigned into two groups: 4 subjects were assigned to the control group, and the other 3 subjects to the treatment group.
+There were six male participants and one female participant.
+Participants were compensated in the amount of 15 dollars for their participation.
+
+All subjects had prior experience with both Java and regular expressions, assessed using a preliminary survey similar to the one described in Section II.
+Of note, the subjects in this study were slightly less skilled with Java and regular expressions than in the previous survey.
+This is consistent with the responses of the PhD students who participated in the online survey, who were also slightly less experienced on average.
+The only other significant difference in responses was that the PhD students were slightly less likely to prefer the use of external tools.
+The difference in the case of regular expressions was 13\%.
+
+\subsection{Hypotheses}
+In designing the regular expression palette, we had hypothesized that users would experience difficulties with two particular aspects of the Java Pattern API: that it used a factory pattern for instantiation, as opposed to the standard instantiation construct \cite{ellis_factory_2007}, and that it required special care with escape sequences -- escape sequences in the pattern must themselves be escaped, because the regular expression is written as a Java string literal.
+We observed difficulties with both of these issues in the control group, but not in the treatment group.
+The treatment group also completed more tasks than the control group on average (7 vs.
+6.) The behaviors observed in these groups are described below.
+
+\subsection{Preliminary Reading}
+All subjects were somewhat rusty on the details of the Java Pattern API.
+After reading the prompt for the first task, all subjects began by searching for and reviewing documentation related to regular expressions.
+In all but one of these cases, the subject looked at the API documentation for the Java Pattern class during this initial review.
+The remaining subject, a member of the control group, referred to quick reference documentation provided by an external tool.
+In all cases, the documentation was left accessible in a browser window throughout the study, often displayed simultaneously alongside the coding window due to the large screen available for the study.
+
 %
@@
 %\subsubsection{Task Structure}
-
-The subjects were given a skeleton class which contained the instruction for how to complete the task and a method stub called \texttt{getRegexPattern} for each task.
-They are asked to fill the body of the \texttt{getRegexPattern} method.
-The \texttt{Task1.java} - \texttt{Task9.java} files were listed in the Eclipse project explorer so that the subjects could easily open and reopen the tasks in any order that they wanted.
 
@@
 
-\section{Pilot Study Results}
-
-\subsection{Participants}
-
-As a pilot study, we recruited 7 PhD students from School of Computer Science at Carnegie Mellon University.
-The subjects were randomly assigned into two groups: 4 subjects were assigned to the control group, and the other 3 subjects to the treatment group.
-There were six male participants and one female participant.
-Participants were compensated in the amount of 15 dollars for their participation.
-
-All participants had prior experience with both Java and regular expressions, but they were all somewhat rusty on the details of the API.
-After reading the prompt for the first task, all subjects began by searching for and reviewing documentation related to regular expressions.
-In all but one of these cases, the subject looked at the API documentation for the Java Pattern class during this initial review.
-The remaining subject, a member of the control group, referred to quick reference documentation provided by an external tool.
-In all cases, the documentation was left accessible in a browser window throughout the study, often displayed simultaneously alongside the coding window due to the large screen available for the study.
-
-\subsection{Control}
+\subsection{Control Group}
 In our previous online survey, we had found that no single strategy for writing regular expressions dominated the others.
-About 40\% of participants indicated that they would use an external tool, 30\% indicated a preference for writing a separate test script, and about 13\% of participants chose the guess-and-check strategy.
-Although our sample size is too small to make a quantitative comparison, we did observe each of these strategies in the control group.
+We observed each of the common strategies -- use of external tools, test scripts and guess-and-check.
 
@@
 After writing a full regular expression and attempting to test it using the tool, the subject became dissatisfied with it and switched to a different tool, regextester.com.
-This tool too appeared to be unsatisfactory, as all subsequent tasks were performed without the aid of external tools or tests of any kind.
+This tool too appeared to be unsatisfactory, as all subsequent tasks were performed without the aid of external tools or tests.
 
@@
 The code could only check one example at a time and the subject used it to check a single positive example per task.
-
-In designing the regular expression palette, we had hypothesized that users would experience difficulties with two particular aspects of the Java Pattern API: that it used a factory pattern for instantiation, as opposed to the standard instantiation construct (new), and that it required special care with escape sequences -- escape sequences in the pattern must themselves be escaped, because the regular expression is written as a Java string literal.
-We observed difficulties with both of these issues in the control group.
 
@@
 Thus, he  spent several minutes looking up how to match whitespace correctly.
-In doing so, he found a description of the double-escape problem in Java and could fix the problem after that.
-\subsection{Treatment}
+In doing so, he found a description of the double-escape problem in Java and fixed the problem after that.
+\subsection{Treatment Group}
 One subject was already familiar with an external tool, regexpal.com (also used by a subject in the control group), and used it variously throughout the task.
@@
 At other times, the subject used the external tool itself.
-The subject indicated that in some cases, he was not sure that our palette was free of bugs (when something unexpectedly matched or did not match\footnote{We observed no actual errors related to regex matching in our tool.}), and also because the external tool provided syntax and substring highlighting, a useful feature for complex regular expressions.
+The subject indicated that in some cases, he was not sure that our palette was free of bugs (when something unexpectedly matched or did not match.
+We did not observe any actual errors related to regex matching in our tool, however.), and also because the external tool provided syntax and substring highlighting, a useful feature for complex regular expressions.
 When done with the external tool, the subject pasted the pattern into our tool to generate the appropriate code.
@@
 The other two subjects in the treatment group did not use any external tools.
-Both subjects decided not to use our palette for the first task (likely because they forgot that it was available, since they spent some time looking up API documentation after our initial demo with the color palette.) They both recognized the need for factory pattern and also both had an initial error related to escaping that they were able to resolve before moving on.
+Both subjects decided not to use our palette for the first task (likely because they forgot that it was available, since they spent some time looking up API documentation after our initial demo with the color palette.) They both recognized the need for the factory pattern and also both had an initial error related to escaping that they were able to resolve before moving on.
 One of these subjects also began to write tests within a \verb|main| method.
@@
 Although we provided a key saying that green meant that ``pattern matches string'', the headings of the two columns: ``[should \verb1|1 should NOT] match:'', may have caused confusion.
-We had changed the meaning of these colors due to feedback from our initial presentation of the tool in class, so it is clear that regardless of the interpretation given, some subjects were confused.
+We had changed the meaning of these colors due to feedback from our initial presentation of the tool, so it is clear that regardless of the interpretation given, some subjects were confused.
 
@@
 None struggled with issues related to escaping and instantiation, for example.
-Although the sample size is not large enough to make conclusive or quantitative judgements, it was observed that the treatment group completed more of the tasks than the test group on average (7 tasks for the treatment group vs.
+Although the sample size is not large enough to make many quantitative judgements, it was observed that the treatment group completed more of the tasks than the test group on average (7 tasks for the treatment group vs.
 6 for the control group).
 All of the subjects in the treatment group, as well as members of the control group who were shown the palette after the study, indicated that they felt that the palette was helpful, and several provided specific suggestions for improvements.
+
 %\section{Future Work}
@@
 %
+
+\subsection{Threats to Validity}
+In addition to the small sample size and the sampling bias discussed earlier, subjects in the treatment group may have been biased to use our features due to their novelty (although two subjects did not use the palette until their second task).
+We only tested a regular expression palette, so other types of palettes may not necessarily be as useful, and this palette had significant flaws that were corrected only after this study.
+
 \section{Related Work}
+In addition to the code completion work discussed in the introduction, some other research areas are related to active code completion. 
 \subsection{Active Libraries}
@@
 Motivated by evidence that integrating highly-specialized tools directly into a developer's workflow is useful, we have developed the concept of active code completion as a generalization of conventional code completion.
-We validated the usefulness and generated a number of use cases and design constraints for this concept with an extensive survey of professional developers.
-Based on these findings, we developed an active code completion architecture that makes several novel design decisions to ease development, deployment and discovery of user-defined palettes.
-We created a palette for regular expression tasks and validated its usefulness with a user study.
-We claim that this provides evidence for the more general claim that integrating palettes into a code editor is useful, and that active code completion systems like Graphite considerably ease this process.
+We validated the usefulness by generating a number of use cases and developed general design constraints for such tools, as well as the underlying architecture, by conducting an extensive survey of professional developers.
+Based on these findings, we developed Graphite, an active code completion architecture that makes several novel design decisions that ease the development, deployment and discovery of user-defined palettes.
+We created palettes for color and regular expression classes and validated the latter palette's usefulness with a pilot study, providing evidence for the more general claim that integrating palettes into code completion is useful.
+We claim that active code completion systems like Graphite will considerably ease this process.
+
+\section{Availability}
+Graphite is free, open-source software available from the URL on the first page. We encourage readers to install it and tell us about any interesting palettes that they develop.
 
@@
 \section*{Acknowledgment}
-% Funding sources
-% Class 
-% PLAID group / Jonathan Aldrich
-% Participants
-% Anonymous reviewers
-% UIUC people
-
-Funding for this work comes in part from the Korea Foundation for Advanced Studies (KFAS) and in part from NSF grant CCF-0811610.
-Any opinions, findings and conclusions or recommendations expressed in this material are those of the author(s) and do not necessarily reflect those of KFAS or the NSF.
+{We would like to thank the participants in our survey and pilot study, Jonathan Aldrich, the PLAID group, the students of 05-899D, anonymous reviewers and the UIUC Software Engineering Seminar group for valuable feedback.
+This material is based upon work supported by the National Science Foundation under Grant No.
+CCF-0811610.
+CO is supported by a DOE CSGF under Grant No.
+DE-FG02-97ER25308, and YY is supported by the Korea Foundation for Advanced Studies.}
 

### 2012-04-12 icse12/graphite-icse12.tex
Margin and font issues for ICSE12 final submission fixed.
Moved graphite URL into a footnote per publisher request.

@@
 Pittsburgh, PA, USA\\
-\small{\texttt{\{comar,youngseok,tlatoza,bam\}$@$cs.cmu.edu}}\\
-\small{\texttt{\url{http://www.cs.cmu.edu/~NatProg/graphite.html}}}}
-}
+\small{{\{comar,youngseok,tlatoza,bam\}$@$cs.cmu.edu}}\\
+}}
 
@@
 \begin{center}
-\includegraphics[width=\textwidth]{color_palette.png}\end{center}
+\includegraphics[width=40pc]{color_palette.png}\end{center}
 \caption{(a) An example code completion palette associated with the \texttt{Color} class. (b) The source code generated by this palette.}
@@
 When selected, the corresponding palette is shown, replacing the standard code completion menu.
-Figure 1a gives an example of a simple palette that may be associated with the \verb|Color| class\footnote{A video demonstrating this process is available from the webpage above.}.
+Figure 1a gives an example of a simple palette that may be associated with the \verb|Color| class\footnote{A video demonstrating this process is available at \\\url{http://www.cs.cmu.edu/~NatProg/graphite.html}.}.
 
@@
 \begin{figure}
+\vspace{2mm}
 \begin{tabular}{lccc}
@@
 \label{strategies}
+\vspace{-2mm}
 \end{figure}
@@
 \begin{figure}\label{collections}
+\vspace{2mm}
 \begin{center}
@@
 \end{center}
+\vspace{-2mm}
 \end{figure}
@@
 \vspace{2mm}
-\includegraphics[scale=.6]{regex.png}\end{center}
+\includegraphics[width=33pc]{regex.png}\end{center}
 \caption{The regular expression palette developed using Graphite and used by subjects in the treatment group of the user study described in Sec. VI.}
@@
 CO is supported by a DOE CSGF under Grant No.
-DE-FG02-97ER25308, and YY is supported by the Korea Foundation for Advanced Studies.}
+DE-FG02-97ER25308 and NSF grant CCF-1116907, and YY is supported by the Korea Foundation for Advanced Studies.}
 

### 2012-07-02 popl13/sigplanconf-template.tex
Added POPL13

%-----------------------------------------------------------------------------
%
%               Template for sigplanconf LaTeX Class
%
% Name:         sigplanconf-template.tex
%
% Purpose:      A template for sigplanconf.cls, which is a LaTeX 2e class
%               file for SIGPLAN conference proceedings.
%
% Author:       Paul C. Anagnostopoulos
%               Windfall Software
%               978 371-2316
%               paul@windfall.com
%
% Created:      15 February 2005
%
%-----------------------------------------------------------------------------

\documentclass{sigplanconf}

% The following \documentclass options may be useful:
%
% 10pt          To set in 10-point type instead of 9-point.
% 11pt          To set in 11-point type instead of 9-point.
% authoryear    To obtain author/year citation style instead of numeric.

\usepackage{amsmath}

\begin{document}

\conferenceinfo{WXYZ '05}{date, City.} 
\copyrightyear{2005} 
\copyrightdata{[to be supplied]}

\titlebanner{banner above paper title}        % These are ignored unless
\preprintfooter{short description of paper}   % 'preprint' option specified.

\title{Type-Determined Compilation Supports Fine-Grained Composition of Language Features}
\subtitle{Removing the Barriers to Adoption of Programming Language Research}

\authorinfo{Cyrus Omar}
           {Computer Science Department\\
           Carnegie Mellon University}
           {cyrus@cmu.edu}
\authorinfo{Jonathan Aldrich}
           {Institute for Software Research\\
           Carnegie Mellon University}
           {jonathan.aldrich@cs.cmu.edu}

\maketitle

\begin{abstract}
Both general-purpose and domain-specific programming languages are commonly distributed as monolithic collections of syntactic and semantic rules useful for a particular set of tasks. However, domain experts and programming language designers typically think in terms of fine-grained composition of "language features", some of which may require compiler support to operate correctly. To support extensible compilation at the language level, we describe a framework which requires type definitions, written in a general purpose definition language, to determine their own data layout, static type checking rules, type inference rules and syntactic transformations for each operation which they support. Because these behaviors are associated with types rather than monolithic language definitions or compiler implementations, they can be composed more easily. We demonstrate how existing language designs -- a polymorphic extension of the statically-typed imperative language OpenCL (based on C99), and example dynamically typed and functional languages -- can be naturally implemented, extended with domain-specific constructs, and interoperate within this framework, without a loss of run-time performance. We also demonstrate that such a system naturally supports staged compilation.
\end{abstract}

\category{CR-number}{subcategory}{third-level}

\terms
term1, term2

\keywords
keyword1, keyword2

\section{Introduction}

A programming language today is a monolithic collection of preferred rules and abstractions, along with a compiler or interpreter to verify and execute programs written using these abstractions. When programming language researchers or domain experts discover new features that may improve the expressiveness, performance, security or correctness of programs, they often turn to tools which allow them to write a new language\cite{abc}. This proliferation of distinct programming languages forces programmers to choose between production-quality languages lacking useful abstractions and prototype languages which may nevertheless offer a small number of useful features or guarantees about performance, security or correctness. Overwhelmingly, programmers have favored mature languages at the expense of these considerations. While useful abstractions occasionally are incorporated into mature languages, this happens slowly and at the whims of the language designers. This hinders the translation and evaluation of innovations in  programming languages research, and prevents domain experts from being able to easily incorporate domain-specific logic into their programming languages, particularly at compile-time and in a performant manner.

\section{System Design}
\subsection{Polymorphism and Type Inference}

\subsection{Type Definitions}

\subsection{Control of Data Layout}

\subsection{Custom Type Checking and Type Inference Rules}

\subsection{Syntactic Transformations and Macros}

\section{Language Implementations}
\subsection{Polymorphic OpenCL}

\subsubsection{Higher Order Functions}

\subsubsection{Performance}

\subsection{Dynamic Languages}

\subsection{Functional Languages}

\section{Domain-Specific Extensions}
\subsection{Scientific Unit Checks}
\subsection{HTML Structure Checking}
\section{Related Work}
\section{Discussion}
%\appendix
%\section{Appendix Title}
%
%This is the text of the appendix, if you need one.

\acks

Acknowledgments, if needed.

% We recommend abbrvnat bibliography style.

\bibliographystyle{abbrvnat}
\bibliography{research}

% The bibliography should be embedded for final submission.

\begin{thebibliography}{}
\softraggedright

\bibitem[Smith et~al.(2009)Smith, Jones]{smith02}
P. Q. Smith, and X. Y. Jones. ...reference text...

\end{thebibliography}

\end{document}

### 2012-08-30 ipdps13/clq-ipdps13.tex
Added initial port of IPDPS 2013 paper.

%% bare_conf.tex
%% V1.3
%% 2007/01/11
%% by Michael Shell
%% See:
%% http://www.michaelshell.org/
%% for current contact information.
%%
%% This is a skeleton file demonstrating the use of IEEEtran.cls
%% (requires IEEEtran.cls version 1.7 or later) with an IEEE conference paper.
%%
%% Support sites:
%% http://www.michaelshell.org/tex/ieeetran/
%% http://www.ctan.org/tex-archive/macros/latex/contrib/IEEEtran/
%% and
%% http://www.ieee.org/

%%*************************************************************************
%% Legal Notice:
%% This code is offered as-is without any warranty either expressed or
%% implied; without even the implied warranty of MERCHANTABILITY or
%% FITNESS FOR A PARTICULAR PURPOSE! 
%% User assumes all risk.
%% In no event shall IEEE or any contributor to this code be liable for
%% any damages or losses, including, but not limited to, incidental,
%% consequential, or any other damages, resulting from the use or misuse
%% of any information contained here.
%%
%% All comments are the opinions of their respective authors and are not
%% necessarily endorsed by the IEEE.
%%
%% This work is distributed under the LaTeX Project Public License (LPPL)
%% ( http://www.latex-project.org/ ) version 1.3, and may be freely used,
%% distributed and modified. A copy of the LPPL, version 1.3, is included
%% in the base LaTeX documentation of all distributions of LaTeX released
%% 2003/12/01 or later.
%% Retain all contribution notices and credits.
%% ** Modified files should be clearly indicated as such, including  **
%% ** renaming them and changing author support contact information. **
%%
%% File list of work: IEEEtran.cls, IEEEtran_HOWTO.pdf, bare_adv.tex,
%%                    bare_conf.tex, bare_jrnl.tex, bare_jrnl_compsoc.tex
%%*************************************************************************

% *** Authors should verify (and, if needed, correct) their LaTeX system  ***
% *** with the testflow diagnostic prior to trusting their LaTeX platform ***
% *** with production work. IEEE's font choices can trigger bugs that do  ***
% *** not appear when using other class files.                            ***
% The testflow support page is at:
% http://www.michaelshell.org/tex/testflow/

% Note that the a4paper option is mainly intended so that authors in
% countries using A4 can easily print to A4 and see how their papers will
% look in print - the typesetting of the document will not typically be
% affected with changes in paper size (but the bottom and side margins will).
% Use the testflow package mentioned above to verify correct handling of
% both paper sizes by the user's LaTeX system.
%
% Also note that the "draftcls" or "draftclsnofoot", not "draft", option
% should be used if it is desired that the figures are to be displayed in
% draft mode.
%
\documentclass[10pt, conference, compsocconf]{IEEEtran}
% Add the compsocconf option for Computer Society conferences.
%
% If IEEEtran.cls has not been installed into the LaTeX system files,
% manually specify the path to it like:
% \documentclass[conference]{../sty/IEEEtran}

% Some very useful LaTeX packages include:
% (uncomment the ones you want to load)

% *** MISC UTILITY PACKAGES ***
%
%\usepackage{ifpdf}
% Heiko Oberdiek's ifpdf.sty is very useful if you need conditional
% compilation based on whether the output is pdf or dvi.
% usage:
% \ifpdf
%   % pdf code
% \else
%   % dvi code
% \fi
% The latest version of ifpdf.sty can be obtained from:
% http://www.ctan.org/tex-archive/macros/latex/contrib/oberdiek/
% Also, note that IEEEtran.cls V1.7 and later provides a builtin
% \ifCLASSINFOpdf conditional that works the same way.
% When switching from latex to pdflatex and vice-versa, the compiler may
% have to be run twice to clear warning/error messages.

% *** CITATION PACKAGES ***
%
%\usepackage{cite}
% cite.sty was written by Donald Arseneau
% V1.6 and later of IEEEtran pre-defines the format of the cite.sty package
% \cite{} output to follow that of IEEE. Loading the cite package will
% result in citation numbers being automatically sorted and properly
% "compressed/ranged". e.g., [1], [9], [2], [7], [5], [6] without using
% cite.sty will become [1], [2], [5]--[7], [9] using cite.sty. cite.sty's
% \cite will automatically add leading space, if needed. Use cite.sty's
% noadjust option (cite.sty V3.8 and later) if you want to turn this off.
% cite.sty is already installed on most LaTeX systems. Be sure and use
% version 4.0 (2003-05-27) and later if using hyperref.sty. cite.sty does
% not currently provide for hyperlinked citations.
% The latest version can be obtained at:
% http://www.ctan.org/tex-archive/macros/latex/contrib/cite/
% The documentation is contained in the cite.sty file itself.

% *** GRAPHICS RELATED PACKAGES ***
%
\ifCLASSINFOpdf
  % \usepackage[pdftex]{graphicx}
  % declare the path(s) where your graphic files are
  % \graphicspath{{../pdf/}{../jpeg/}}
  % and their extensions so you won't have to specify these with
  % every instance of \includegraphics
  % \DeclareGraphicsExtensions{.pdf,.jpeg,.png}
\else
  % or other class option (dvipsone, dvipdf, if not using dvips). graphicx
  % will default to the driver specified in the system graphics.cfg if no
  % driver is specified.
  % \usepackage[dvips]{graphicx}
  % declare the path(s) where your graphic files are
  % \graphicspath{{../eps/}}
  % and their extensions so you won't have to specify these with
  % every instance of \includegraphics
  % \DeclareGraphicsExtensions{.eps}
\fi
% graphicx was written by David Carlisle and Sebastian Rahtz. It is
% required if you want graphics, photos, etc. graphicx.sty is already
% installed on most LaTeX systems. The latest version and documentation can
% be obtained at: 
% http://www.ctan.org/tex-archive/macros/latex/required/graphics/
% Another good source of documentation is "Using Imported Graphics in
% LaTeX2e" by Keith Reckdahl which can be found as epslatex.ps or
% epslatex.pdf at: http://www.ctan.org/tex-archive/info/
%
% latex, and pdflatex in dvi mode, support graphics in encapsulated
% postscript (.eps) format. pdflatex in pdf mode supports graphics
% in .pdf, .jpeg, .png and .mps (metapost) formats. Users should ensure
% that all non-photo figures use a vector format (.eps, .pdf, .mps) and
% not a bitmapped formats (.jpeg, .png). IEEE frowns on bitmapped formats
% which can result in "jaggedy"/blurry rendering of lines and letters as
% well as large increases in file sizes.
%
% You can find documentation about the pdfTeX application at:
% http://www.tug.org/applications/pdftex

% *** MATH PACKAGES ***
%
%\usepackage[cmex10]{amsmath}
% A popular package from the American Mathematical Society that provides
% many useful and powerful commands for dealing with mathematics. If using
% it, be sure to load this package with the cmex10 option to ensure that
% only type 1 fonts will utilized at all point sizes. Without this option,
% it is possible that some math symbols, particularly those within
% footnotes, will be rendered in bitmap form which will result in a
% document that can not be IEEE Xplore compliant!
%
% Also, note that the amsmath package sets \interdisplaylinepenalty to 10000
% thus preventing page breaks from occurring within multiline equations. Use:
%\interdisplaylinepenalty=2500
% after loading amsmath to restore such page breaks as IEEEtran.cls normally
% does. amsmath.sty is already installed on most LaTeX systems. The latest
% version and documentation can be obtained at:
% http://www.ctan.org/tex-archive/macros/latex/required/amslatex/math/

% *** SPECIALIZED LIST PACKAGES ***
%
%\usepackage{algorithmic}
% algorithmic.sty was written by Peter Williams and Rogerio Brito.
% This package provides an algorithmic environment fo describing algorithms.
% You can use the algorithmic environment in-text or within a figure
% environment to provide for a floating algorithm. Do NOT use the algorithm
% floating environment provided by algorithm.sty (by the same authors) or
% algorithm2e.sty (by Christophe Fiorio) as IEEE does not use dedicated
% algorithm float types and packages that provide these will not provide
% correct IEEE style captions. The latest version and documentation of
% algorithmic.sty can be obtained at:
% http://www.ctan.org/tex-archive/macros/latex/contrib/algorithms/
% There is also a support site at:
% http://algorithms.berlios.de/index.html
% Also of interest may be the (relatively newer and more customizable)
% algorithmicx.sty package by Szasz Janos:
% http://www.ctan.org/tex-archive/macros/latex/contrib/algorithmicx/

% *** ALIGNMENT PACKAGES ***
%
%\usepackage{array}
% Frank Mittelbach's and David Carlisle's array.sty patches and improves
% the standard LaTeX2e array and tabular environments to provide better
% appearance and additional user controls. As the default LaTeX2e table
% generation code is lacking to the point of almost being broken with
% respect to the quality of the end results, all users are strongly
% advised to use an enhanced (at the very least that provided by array.sty)
% set of table tools. array.sty is already installed on most systems. The
% latest version and documentation can be obtained at:
% http://www.ctan.org/tex-archive/macros/latex/required/tools/

%\usepackage{mdwmath}
%\usepackage{mdwtab}
% Also highly recommended is Mark Wooding's extremely powerful MDW tools,
% especially mdwmath.sty and mdwtab.sty which are used to format equations
% and tables, respectively. The MDWtools set is already installed on most
% LaTeX systems. The lastest version and documentation is available at:
% http://www.ctan.org/tex-archive/macros/latex/contrib/mdwtools/

% IEEEtran contains the IEEEeqnarray family of commands that can be used to
% generate multiline equations as well as matrices, tables, etc., of high
% quality.

%\usepackage{eqparbox}
% Also of notable interest is Scott Pakin's eqparbox package for creating
% (automatically sized) equal width boxes - aka "natural width parboxes".
% Available at:
% http://www.ctan.org/tex-archive/macros/latex/contrib/eqparbox/

% *** SUBFIGURE PACKAGES ***
%\usepackage[tight,footnotesize]{subfigure}
% subfigure.sty was written by Steven Douglas Cochran. This package makes it
% easy to put subfigures in your figures. e.g., "Figure 1a and 1b". For IEEE
% work, it is a good idea to load it with the tight package option to reduce
% the amount of white space around the subfigures. subfigure.sty is already
% installed on most LaTeX systems. The latest version and documentation can
% be obtained at:
% http://www.ctan.org/tex-archive/obsolete/macros/latex/contrib/subfigure/
% subfigure.sty has been superceeded by subfig.sty.

%\usepackage[caption=false]{caption}
%\usepackage[font=footnotesize]{subfig}
% subfig.sty, also written by Steven Douglas Cochran, is the modern
% replacement for subfigure.sty. However, subfig.sty requires and
% automatically loads Axel Sommerfeldt's caption.sty which will override
% IEEEtran.cls handling of captions and this will result in nonIEEE style
% figure/table captions. To prevent this problem, be sure and preload
% caption.sty with its "caption=false" package option. This is will preserve
% IEEEtran.cls handing of captions. Version 1.3 (2005/06/28) and later 
% (recommended due to many improvements over 1.2) of subfig.sty supports
% the caption=false option directly:
%\usepackage[caption=false,font=footnotesize]{subfig}
%
% The latest version and documentation can be obtained at:
% http://www.ctan.org/tex-archive/macros/latex/contrib/subfig/
% The latest version and documentation of caption.sty can be obtained at:
% http://www.ctan.org/tex-archive/macros/latex/contrib/caption/

% *** FLOAT PACKAGES ***
%
%\usepackage{fixltx2e}
% fixltx2e, the successor to the earlier fix2col.sty, was written by
% Frank Mittelbach and David Carlisle. This package corrects a few problems
% in the LaTeX2e kernel, the most notable of which is that in current
% LaTeX2e releases, the ordering of single and double column floats is not
% guaranteed to be preserved. Thus, an unpatched LaTeX2e can allow a
% single column figure to be placed prior to an earlier double column
% figure. The latest version and documentation can be found at:
% http://www.ctan.org/tex-archive/macros/latex/base/

%\usepackage{stfloats}
% stfloats.sty was written by Sigitas Tolusis. This package gives LaTeX2e
% the ability to do double column floats at the bottom of the page as well
% as the top. (e.g., "\begin{figure*}[!b]" is not normally possible in
% LaTeX2e). It also provides a command:
%\fnbelowfloat
% to enable the placement of footnotes below bottom floats (the standard
% LaTeX2e kernel puts them above bottom floats). This is an invasive package
% which rewrites many portions of the LaTeX2e float routines. It may not work
% with other packages that modify the LaTeX2e float routines. The latest
% version and documentation can be obtained at:
% http://www.ctan.org/tex-archive/macros/latex/contrib/sttools/
% Documentation is contained in the stfloats.sty comments as well as in the
% presfull.pdf file. Do not use the stfloats baselinefloat ability as IEEE
% does not allow \baselineskip to stretch. Authors submitting work to the
% IEEE should note that IEEE rarely uses double column equations and
% that authors should try to avoid such use. Do not be tempted to use the
% cuted.sty or midfloat.sty packages (also by Sigitas Tolusis) as IEEE does
% not format its papers in such ways.

% *** PDF, URL AND HYPERLINK PACKAGES ***
%
%\usepackage{url}
% url.sty was written by Donald Arseneau. It provides better support for
% handling and breaking URLs. url.sty is already installed on most LaTeX
% systems. The latest version can be obtained at:
% http://www.ctan.org/tex-archive/macros/latex/contrib/misc/
% Read the url.sty source comments for usage information. Basically,
% \url{my_url_here}.

% *** Do not adjust lengths that control margins, column widths, etc. ***
% *** Do not use packages that alter fonts (such as pslatex).         ***
% There should be no need to do such things with IEEEtran.cls V1.6 and later.
% (Unless specifically asked to do so by the journal or conference you plan
% to submit to, of course. )

% correct bad hyphenation here
\hyphenation{op-tical net-works semi-conduc-tor}
\input{macros}

\begin{document}
%
% paper title
% can use linebreaks \\ within to get better formatting as desired
\title{\Ace: An Actively-Typed Compilation Environment\\for High-Performance Computing}

% author names and affiliations
% use a multiple column layout for up to two different
% affiliations

\author{\IEEEauthorblockN{Cyrus Omar, Jonathan Aldrich}
\IEEEauthorblockA{School of Computer Science\\
Carnegie Mellon University\\
Pittsburgh, PA, USA\\
\texttt{\{comar,jonathan.aldrich\}@cs.cmu.edu}}}

% conference papers do not typically use \thanks and this command
% is locked out in conference mode. If really needed, such as for
% the acknowledgment of grants, issue a \IEEEoverridecommandlockouts
% after \documentclass

% for over three affiliations, or if they all won't fit within the width
% of the page, use this alternative format:
% 
%\author{\IEEEauthorblockN{Michael Shell\IEEEauthorrefmark{1},
%Homer Simpson\IEEEauthorrefmark{2},
%James Kirk\IEEEauthorrefmark{3}, 
%Montgomery Scott\IEEEauthorrefmark{3} and
%Eldon Tyrell\IEEEauthorrefmark{4}}
%\IEEEauthorblockA{\IEEEauthorrefmark{1}School of Electrical and Computer Engineering\\
%Georgia Institute of Technology,
%Atlanta, Georgia 30332--0250\\ Email: see http://www.michaelshell.org/contact.html}
%\IEEEauthorblockA{\IEEEauthorrefmark{2}Twentieth Century Fox, Springfield, USA\\
%Email: homer@thesimpsons.com}
%\IEEEauthorblockA{\IEEEauthorrefmark{3}Starfleet Academy, San Francisco, California 96678-2391\\
%Telephone: (800) 555--1212, Fax: (888) 555--1212}
%\IEEEauthorblockA{\IEEEauthorrefmark{4}Tyrell Inc., 123 Replicant Street, Los Angeles, California 90210--4321}}

% use for special paper notices
%\IEEEspecialpapernotice{(Invited Paper)}

% make the title area
\maketitle

\begin{abstract}
In this paper, we introduce the Ace programming language and demonstrate its practicality as a foundational tool for both high-performance computing researchers and practitioners. Ace is a statically-typed language that shares a syntax with Python and uses the Python language itself as a multi-purpose compile-time metalanguage. Ace has a minimal core that delegates control over many aspects of compilation to user modules via a novel extension mechanism called {\em active type-checking and translation (\ATT)}. Specifically, users specify new primitive types and operations by equipping type definitions -- first-class objects in the metalanguage -- with compile-time methods. The compiler invokes these methods when type checking and translating expressions, according to a fixed type dispatch protocol associated with each syntactic form.

\end{abstract}

%\begin{IEEEkeywords}
%programming languages, heterogeneous computing, metaprogramming, code generation

%\end{IEEEkeywords}

% For peer review papers, you can put extra information on the cover
% page as needed:
% \ifCLASSOPTIONpeerreview
% \begin{center} \bfseries EDICS Category: 3-BBND \end{center}
% \fi
%
% For peerreview papers, this IEEEtran command inserts a page break and
% creates the second title. It will be ignored for other modes.
\IEEEpeerreviewmaketitle

\section{Introduction}
Researchers design and implement novel parallel programming abstractions\- on a regular basis. Each abstraction aims to make some class of developers more productive by making appropriate trade-offs between performance, portability, verifiability and ease-of-use. It can be observed that the most widely-adopted abstractions to date are those that been implemented as libraries that can be easily deployed within existing, widely-used languages. Abstractions that require introducing new primitive constructs, changing the semantics of existing constructs in particular contexts, or controlling compilation in a fine-grained manner cannot take this approach. This has led to a proliferation of specialized programming languages, each designed around a few privileged abstractions. Unfortunately, such languages have not been widely adopted by practitioners.

Computer-aided simulations and data analysis techniques have transformed science and engineering. Surveys show that scientists and engineers now spend up to 40\% of their time writing software, generally individually or in small groups \cite{howison2011scientific}\cite{hannay2009scientists}. Most of this software targets conventional desktop hardware, but about 20\% of scientists also target either local clusters or super\-computers for more numerically-intensive computations \cite{hannay2009scientists}. In recent years, GPUs and other kinds of coprocessors have emerged as powerful platforms for high-performance computing as well.

Many scientists and engineers prefer high-level languages like MATLAB, Python, R and Perl \cite{nguyen2010survey}, due to their powerful domain-specific libraries and their focus on productivity and flexibility. These languages are typically interpreted rather than compiled and generally do not feature static type systems, relying instead on run-time (``dynamic'') type lookup and indirection. Although this can increase the flexibility of the language, these features also negatively impact  performance. Along code paths where performance appears to be a bottleneck, developers typically turn to traditional statically-typed low-level languages like C and Fortran \cite{4222616}. These languages require explicit type annotations on variables and give users explicit control over data layout and heap allocation. Although this makes writing programs more tedious and error-prone, it can also significantly  improve performance, particularly if attributes of the target architecture (e.g. caching behavior) are considered.

While a sequential reimplementation of a critical code path using a low-level language will produce a modest speedup, more significant speedups on modern hardware require parallelizing over many processor cores and, on massively-parallel machines, many interconnected nodes. Despite well-known difficulties, low-level approaches that use a form of shared-memory  multithreading (e.g. pthreads, OpenMP) paired with explicit message passing between nodes (typically using MPI) remain widely used \cite{4222616}\cite{basili2008understanding}.

Although researchers often propose higher-level language features and parallel programming abstractions that aim to strike a better balance between raw performance, productivity, code portability and verifiability (see Section 2), end-users remain skeptical of new approaches. This viewpoint was perhaps most succinctly expressed by a participant interviewed in a recent study \cite{basili2008understanding}, who stated ``I hate MPI, I hate C++. [But] if I had to choose again, I would probably choose the same.'' Although this sentiment is easy to dismiss as paradoxical, we believe that it demands direct examination by those in the research community working to advance the practice of scientific and high-performance computing. We begin in Section 2 by developing a set of design and adoption criteria for new languages and abstractions based on prior empirical studies of these {\it professional end-user developers} as well as our observations of characteristics common to successful projects in the past.

%
% Two types of languages see widespread use: productivity-oriented and performance-oriented
% Significant diversity in productivity-oriented languages. Valued for clean syntax and high-level abstractions. Usually very slow. Have a FFI to call into a performance-oriented language.
% Performance-oriented languages -- C or Fortran and their derivatives. Direct memory access, low-level abstractions, more verbose. Portable (though often need to include platform-specific optimizations, which can lead to messy code). Not easily extended with new abstractions and primitives (witness: lots of derivative languages that are not composable because no way to implement these abstractions cleanly)
% Generally don't support verifiability very well (performance-oriented languages have simple type systems); only exception is functional languages.
%

% Simplicity (Errors, Syntax, Core Semantics)
% Extensibility (New Abstractions, Domain-Specific Things, Domain- and Platform-Specific Optimizations)
% Performance (Need to allow control over data layout and movement, perhaps optionally)
% Portability (Architectures and operating systems; support new optimizations)
% Backwards-Compatibility (Existing codebases)
% Verifiability (type systems, metaprogramming-based, formal semantics, small TCB)
% Tool Support (syntax highlighting, editors, debuggers, etc.)
% Social Proof (Will not disappear - open source, significant projects have been developed, community)

We then introduce a new programming language named $\cloquence$ (pronounced ``C eloquence'') in Section 3. $\cloquence$ initially targets code paths where low-level languages, particularly OpenCL, would be used today. By employing a collection of powerful front-end language and compiler design techniques, making pragmatic design choices, and building on top of established infrastructure, we argue that $\cloquence$ may uniquely satisfy many of the criteria established in Section 2, particularly those related to ease of use, familiarity, performance and tool support.

Unlike many existing languages, $\cloquence$ is designed to be highly extensible from inception, employing a powerful compile-time language extension and code generation mechanism based around the concept of {\it active libraries} \cite{activelibraries}. We argue that this feature may make $\cloquence$ particularly suitable as a platform for researchers developing new parallel abstractions and other specialized, high-level language features, as well as  development tools.

In Section 4 we briefly describe a case study where the language was used to conduct large-scale spiking neural circuit simulations on GPU hardware, demonstrating the feasibility of this approach on a non-trivial problem. Finally, conclude in Section 6 with future directions for both the $\cloquence$ language and the research community.

\section{Design and Adoption Criteria}
The primary purpose of this paper is to describe the design of a programming language, $\cloquence$. Unfortunately, as in many areas of design, it can be difficult to objectively evaluate the merit of such efforts. To better support such evaluations, researchers in design disciplines typically develop a set of high-level  design criteria that serve as a rubric for evaluating and guiding their design efforts. In programming language design, particularly for scientific and high-performance computing, there have been few concerted efforts to develop a coherent set of design criteria that capture the needs of the targeted developer communities. Similarly, there have been few treatments of adoption criteria for new languages and abstractions in practice, an important issue given the slow rates of adoption today. We have purposefully separated these criteria from the specific language we describe in Section 3 in the hopes of starting a productive discussion within the community.

\subsection{Professional End-User Developers}
An important first step is to identify the developer communities targeted by our design efforts. Researchers in software engineering typically distinguish between {\it end-user developers} and {\it professional developers}. End-user developers have little formal training relevant to the tools that they use, and the tools themselves are relatively simple (spreadsheets, for example). Professional developers, on the other hand, have explicit training or extensive experience with software engineering and software development techniques. Scientists and engineers do not cleanly fall into either of these groups, however -- although formal training in software development is rare \cite{oai:open.ac.uk.OAI2:17673}, members of these groups are more technically literate and demanding than typical end-users. For this reason, researchers have proposed a third group, {\it professional end-user developers}, to describe ``people working in highly technical, knowledge rich professions, such as financial mathematicians, scientists and engineers, who develop their own software in order to advance their own professional goals. \cite{segal2007some}'' 
As such, we consider the literature on novice and end-user programming (cf. \cite{pane1996usability}) in addition to that on programming language design for professional developers to justify the criteria we develop below. Language usability and adoption are not commonly studied topics, so many of the hypotheses we present below also rely on informal observations about languages and tools that have been successful in the past.

\subsection{Design Criteria}
We define design criteria as aspects of a language's design, rather than its implementation, that help developers express their intent naturally and correctly.

\subsubsection{Concise, Familiar and Readable Syntax}\label{syntax}
Despite a considerable amount of evidence pointing toward the value of a well-designed syntax, particularly for end-users in specialized domains \cite{pane1996usability}, the issue is sometimes marginalized. It is likely the case, however, that some libraries and languages experience low adoption due in  part due to the use of a verbose, unfamiliar or unreadable syntactic style.

A simple and concise syntax is common to most of the high-level languages that are widely used in scientific computing. Cordy identifies the principle of conciseness with elimination of redundancy and the availability of reasonable defaults \cite{cordy1992hints}. The high-level languages listed in the introduction have all either made optional, or removed entirely, much of the syntactic overhead characteristic of low-level languages, such as explicit variable declarations and extensive headers or preambles that contain information that can be inferred from the body of the program in most cases. They also typically feature simple array indexing syntax and concise literal forms for common data structures like arrays, sets and maps. A concise and minimal syntax eliminates unnecessary keystrokes and keeps more code visible on screen at a time. Studies have shown that there may be a correlation between lines of code entered and overall error rate, independent of other factors \cite{el2001confounding}.

However, one can also go too far, so these benefits must be balanced with concerns about readability and familiarity. A number of principles have been proposed to operationalize the notion of code readability. The most widely-used collection of principles are Green's cognitive dimensions \cite{green1996usability}. Of particular relevance is the notion of self-consistency, which serves to ensure that similar forms have similar meaning and that there are few subtle or context-dependent distinctions that users must be mindful of. Another important principle has been called {\it closeness of mapping}, expressing the value of a close correspondence between mental models and the formal model as expressed concretely using the language.

Most researchers become familiar with formal notation by studying mathematics. In nearly all commonly used languages in scientific computing, the common mathematical notation is used whenever possible. In contrast, many academic languages have settled on alternative notational styles. For example, the LISP family of languages uses a highly uniform list-based notation, while most functional languages typically borrow notation for function invocation from the lambda calculus, using the form \verb|f x| rather than \verb|f(x)| for function invocation. Although both of these styles have certain benefits, they can impose mental burdens on users who continue to mentally translate them into more familiar notation \cite{anderson1985novice}. Although these and related difficulties can decrease with experience, they remain an important barrier for new users to these languages.

Syntactic cues like whitespace and typography that do not have a formal meaning but rather exist to assist developers are called secondary notation \cite{green1990programming}. Most languages are whitespace-insensitive, while others (notably, Python) enforce consistent uses of whitespace, both in pursuit of consistency and as a technique to eliminate the need for block delimiters. Additionally, a few languages, such as Mathematica, have support for more typographically-rich mathematical notation via a structural editing interface. These techniques appear to be helpful, although we do not know of formal studies that provide evidence for this claim.

\subsubsection{Support for Multiple Paradigms and Abstractions}\label{multiparadigm}
Although the difficulties of low-level and parallel programming are widely acknowledged, there is little consensus on which high-level abstractions are most appropriate for easing this burden. Indeed, it appears likely that no one abstraction will emerge triumphant over the others. Although library-based abstractions are useful, they often suffer from limitations of the language, particularly if they require compile-time support. Primitive language support for a parallel abstraction has been shown to be more usable than a library-based implementation in at least one case \cite{cave2010comparing}. 
Languages that have been designed specifically to explore a single abstraction as a core language feature often see more limited adoption than library-based implementations because they remain difficult to use in circumstances for which an alternative abstraction is more appropriate. Examples of abstractions where language-based implementations remain less frequently used than library-based implementations include the actor model (e.g. Erlang), software transactional memory \cite{harris2003language}, and global arrays (e.g. UPC). As such, it is important that a language support several modes of operation, ideally without excessively favoring one over another.

\subsubsection{Extensibility}\label{extensions}

Although support for multiple paradigms and primitive abstractions can be built into a language design, this leaves control in the hands of the language designer. Novel or domain-specific constructs that may be useful to a small number of users or in rare situations can be difficult to develop and distribute for this reason, leading to the proliferation of new languages as  described above. Language extensibility mechanisms support these use cases by giving users the ability to develop new abstractions and constructs that behave as if they were primitive constructs. If a mechanism is powerful enough, nearly all language constructs may be implemented using it, greatly simplifying the core semantics of a language.

Dynamic languages commonly rely on mechanisms like operator overloading and metaobject protocols \cite{Kiczales91} to provide extensibility via indirection. For example, the Python language allows objects to overload nearly every operator and as well as operations like attribute lookup (\verb|obj.attr|) and assignment. This mechanism is used by a number of libraries to create a more natural interface to a low-level API (e.g. \verb|pycuda|). More open-ended dynamic mechanisms, such as programmatic macros, have also been widely studied but as of yet have seen little adoption in languages used by professional end-users.

Language extensibility for statically-typed languages, on the other hand, remains an active research area. Compile-time metaprogramming systems, which are related to programmatic macros, have been developed for a number of languages (e.g. Template Haskell \cite{sheard2002template}) but these too have seen relatively limited development or adoption thus far. In Section 3, we introduce a novel static language extension mechanism that uses type-based dispatch rather than direct manipulation of syntax trees.

Some researchers have advocated the use of compiler extension mechanisms, rather than mechanisms built into a language itself (cf. \cite{clements2008comparison}). Modern compilers now offer some support for front-end language extensions, although often these  can be quite difficult to use.  Although potentially powerful, this approach can also lead to issues when extensions are not easily composable or when multiple compilers exist for a language. Back-end extensions (that is, extensions that preserve the semantics of the language, improving only performance) are better supported in modern compilers.

Domain-specific language frameworks are a related approach that can serve many of the same goals as language-based  extension mechanisms \cite{fowler2010domain}. These tools ease the development of ``little languages'' and allow for the development and distribution of language features as modules. Domain experts use these tools to develop highly specialized languages that capture domain requirements precisely and allow for natural and concise specifications of scientific models and other structures. A major issue with this approach arises at language boundaries -- interoperability between different domain-specific languages is difficult due to feature mismatches. Another issue is that domain-specific languages often outgrow their initial implementations and begin to need increasingly powerful general-purpose features.

All extensibility mechanisms must be used carefully. Indeed, inexperienced developers can abuse mechanisms like operator overloading to create inscrutable interfaces that cause more problems than they solve. Some languages (notably Java) have taken up the philosophy that even simple language extension mechanisms should not be in the hands of end-users for this reason. Although this continues to be debated, we argue that extensibility is crucial for parallel and scientific programming in particular due to the significant levels of ongoing research into new parallel abstractions and the diverse set of other domain-specific use cases.

\subsubsection{Support for Higher-Order Constructs}\label{hof}
A number of parallel programming data structures and algorithms are of higher-order, meaning that they take other functions or types as arguments or parameters. We argue that support for higher-order programming is critical to language usability in our target domains.

Well-known examples of functions that operate using other functions include fundamental parallel primitives like map, reduce and scan. Fortunately, most modern languages now support using functions as values. Languages like C and Fortran implement this using function pointers, although at considerable syntactic expense. However, unlike functional languages and most dynamic languages, they do not allow anonymous functions (that is, functions that are defined as inline expressions rather than statements), nor functions that close over variables in the surrounding scope. The most recent revision of C++ now has support for closures and closures are also being considered for a future revision of Java. OpenCL does not come with any support for higher-order functions, an issue we explore further in Section 3.

Functions and types that are parameterized by types are often referred to as {\it polymorphic} or {\it generic}. C++ supports this using its template system. Java, ML and Haskell support a simpler form of {\it parametric polymorphism}, and Haskell also supports a more flexible {\it type class} system. C++, Java and Fortran also support {\it function overloading}, allowing multiple versions of a function that differ only according to their argument types. C and OpenCL do not support any form of polymorphism or user-defined function overloading.

Dynamic languages do not require that variables be assigned a type, so generic functions are written using run-time checks that ensure that a particular value supports the specific interface that a function expects. This is sometimes known as ``duck typing'' and is generally considered as a useful feature, due to its considerable flexibility. A promising approach that resembles duck typing, while operating statically, is known as {\it structural typing} \cite{malayeri2009structural}.

\subsubsection{Modularity and Packaging}\label{modularity}
Modularity is a broad term that refers to mechanisms useful for combining and reusing independently developed libraries of code. Languages with good support for modularity promote information hiding, thus localizing the effect of code changes, and allow modules to communicate over well-defined interfaces. There remains widespread disagreement and considerable ongoing research on language support for modularity. Object-oriented methodologies, although common in industrial projects, are used less frequently in scientific codes due to a perceived loss of control and performance \cite{basili2008understanding}. Many dynamic languages support modularity only implicitly using duck typing. Functional languages, on the other hand, often have module systems that enforce correct usage of an interface at compile-time, albeit at some notational cost \cite{tapl}.

The packaging and linking mechanisms available in a language can also significantly impact its usability. Languages like C and C++ use a fragile preprocessor-based packaging mechanism and require separate header files, which often leads to subtle errors and compilation inefficiencies. More recent high-level languages have developed a varied set of mechanisms that are significantly simpler.

\subsubsection{Verifiability}\label{verifiability}
Verifying that a program does not contain errors and that it will operate according to specification (if a specification exists, which can be rare in this domain \cite{oai:open.ac.uk.OAI2:17673}) is a critical concern across all areas of software development. Errors in scientific programs may arise due to problems in basic program logic, as in other domains, but also due to the accumulation of  numerical approximation errors or by violation of domain-specific constraints (e.g. inconsistent scientific units.)

A number of design decisions can influence the difficulty of formal program verification. Broadly stated, unconstrained support for dynamic indirection and direct access to memory have made program verification more difficult in many commonly-used languages. As a result, many classes of errors are only caught at run-time, often due to edge cases that are difficult to test for.

Advanced type systems, matched with appropriately constrained data structures and control constructs, can dramatically increase the likelihood that an error is found at compile-time and even eliminate entire classes of errors \cite{tapl}. A large portion of academic research on programming language design has focused on designing such type systems. Unfortunately, as with many new parallel abstractions, these generally require that a language be modified with new primitives and most work has either been with prototype languages or functional languages like Ocaml, Haskell or Coq, which, due to some of the factors mentioned here, have seen  limited (though growing) adoption in scientific computing so far.

In the absence of a formal proof of correctness, users must test programs with specific inputs, relying both on language-provided run-time checks and user-provided run-time assertions to catch errors. Most languages have unit testing frameworks designed to minimize the burden of developing unit tests, though these are used infrequently in science \cite{oai:open.ac.uk.OAI2:17673}\cite{hannay2009scientists}. Several languages (e.g. Eiffel) also have support for specifying pre- and post-conditions for functions, so that the assertions that must hold about arguments are visible in the function signature, but this feature has not been well-supported by widely-used languages to date.

\subsection{Adoption Criteria}
While most of the design criteria described in the previous sections have been the topic of significant (and ongoing) research, there are a number of other, more practical issues that can significantly affect adoption of a new language or tool. It should be noted that certain design decisions may make it easier to satisfy these adoption to criteria as well, an important consideration given that incentive structures in academia often do not reward efforts to improve a language along these dimensions \cite{howison2011scientific}. We give examples of such design decisions when we describe $\cloquence$ in Section 3.

\subsubsection{Performance}
Performance is, of course, a critical concern in scientific and (particularly \cite{basili2008understanding}) high-performance computing. Low-level languages typically elect to give the user direct access to hardware primitives. High-level languages must rely on a sophisticated compiler to translate high-level abstractions into performant code. Although the latter approach would be ideal, program optimization remains an active research area with many open problems. Indeed, many relevant algorithms have been shown to be NP-Complete, so compilers must rely on heuristics. Humans remain better at inventing and applying heuristics, given enough motivation and experience, than computers in many cases. Indeed, even when the compiler can sometimes produce better optimizations, users generally insist on being able to form a mental model of what their code is doing at the machine-level so that they can {\it reason} about the effects of code changes on overall performance \cite{squires2005programmers}. As such, users generally demand that low-level abstractions remain available alongside high-level abstractions. It may also be useful to support a model where programmers can formalize, package and directly apply optimization heuristics programmatically, rather than relying on a ``black-box'' compiler.

% Squires et al, 2005 -- programmers want to know what is happening under the hood so they can reason about optimization

\subsubsection{Portability}\label{portability}
A number of surveys have revealed that portability is one of the most important issues considered by scientists and engineers \cite{hannay2009scientists}\cite{basili2008understanding}. Several processor architectures and operating systems are in widespread use, with more appearing on a fairly regular basis. Low-level languages like C have generally achieved a reasonable level of portability, although aspects of the language that are underspecified can be sources of errors. The OpenCL language was designed with portability as a major design criteria, and compilers for OpenCL are now available for a diverse collection of processor and accelerator architectures. High-level languages are generally highly portable.

It should be noted, however, that simple portability is not entirely sufficient -- users want to be able to write programs that can be ported to new architectures and operating systems and achieve high performance without significant retuning. This remains an area of active research. Recent efforts to build abstractions that generate efficient code for devices with multi-level memories, as implemented in the Sequoia language \cite{fatahalian2006sequoia}, have  progressed toward this goal.

\subsubsection{Useful Error Messages}\label{errors}
When the compiler or run-time system of a language generates an error, users must determine the source of the problem using the information provided in an error message. Studies have shown that good error messages can dramatically reduce the time it takes to debug programs \cite{marceau2011measuring}. Indeed, a widely-reported frustration with C++ is that it produces overly verbose and cryptic error messages, particularly when using templates.

\subsubsection{Tool and Infrastructure Support}\label{tools}
% Squires 2005 might have a ref
Modern software development now relies on a number of tools to assist with common tasks, such as debuggers,  profilers, syntax-aware editors, documentation generators, style checkers and interactive interpreters. Similarly, most established languages benefit from a centralized packaged repository (e.g. PyPI) and a package manager that can automate the installation of new packages. Such support is not trivial to develop from scratch, but few early adopters are likely willing to do without these tools for non-trivial projects \cite{squires2005programmers}.

\subsubsection{Backwards Compatibility}
End-user communities have produced a number of highly tuned and tested packages that are widely used in their fields. Porting these libraries to a new language requires significant effort and few communities will be willing to do so, particularly before a language is very well established. There is generally a greater willingness to develop wrappers that invoke functions from these packages, however. A powerful foreign function interface, ideally able to handle native code as well as code in existing widely-used high-level languages, enables these efforts to proceed smoothly and can provide a language with a large package library relatively early in its development.

Many of the reasonably successful approaches to date are built incrementally on existing languages as libraries or as simple  extensions to existing languages (e.g. OpenMP, UPC, Co-array Fortran, Cilk++, MPI, CUDA, OpenCL). Notably, this allows existing programs to continue to function correctly, while enabling the gradual integration of the new constructs in parts of the code that would derive the greatest benefit. In contrast, approaches that have required total rewrites have had more difficulty recruiting early adopters (e.g. X10 \cite{charles2005x10}.)

\subsubsection{Learning Material}
Existing languages benefit from a large collection of books, presentations and tutorials that allow new users to get started quickly. The most mature languages benefit further from the availability of courses and professional training seminars. New languages often suffer due to the lack of such polished learning material, or in some cases, due to the lack of {\it any} significant learning material targeted at end users (rather than other language designers.)

\subsubsection{Open Availability}
New languages and abstractions are often viewed with suspicion if they do not come with source code that can be modified under a free software license. The most unencumbered licenses (other than public domain releases) are BSD-style licenses. So called {\it copyleft} licenses like the GPL are also popular, requiring primarily that modifications to the compiler be distributed under the same license.

\subsubsection{Social Proof}\label{social}
Finally, users look to the language's user community for evidence that the language is a serious effort that will not be abandoned and that there are other developers building libraries and sharing information over established communication channels. The availability of non-trivial demonstrations have been cited in surveys as critical to adoption as well \cite{basili2008understanding}. These criteria, as with many of the others described above, can be difficult to satisfy, particularly for ``clean-slate'' language designs, which may help to explain why language adoption is often slow. However, even projects with a narrower scope, such as new parallel or domain-specific abstractions, must tackle these issues.

% What Do Programmers of Parallel Machines Need?
% Squiers et al, 2005
% - Global view
% - Tool support
% - Understanding of what the code is doing
% 
%As should be clear, there are many challenges facing any new language design effort. Many of these challenges apply also to efforts with a narrower scope (e.g. a new parallel or domain-specific abstraction). Although the research community continues to tackle open issues described as design criteria, the challenges described as adoption criteria above have not generally received as much attention.

\section{The $\cloquence$ Language}
We now describe a programming language designed with many of the criteria described in the previous sections in mind, $\cloquence$. $\cloquence$ is built around a minimal core and nearly every construct in the language is implemented as a library. Rather than beginning directly with this extension mechanism, however, let us begin with a concrete example of the language in use.

\subsection{Example: Higher-Order Map for OpenCL}
Figure \ref{map} shows the standard data-parallel map function written using an extension that implements the full OpenCL language as a library. Each thread, indexed by \verb|gid|, applies the transfer function, \verb|fn|, to the  corresponding element of the input array, \verb|in|, writing the result into the corresponding location in the output array, \verb|out|.

\subsection{Syntax}
Readers familiar with the Python programming language will recognize the style of syntax used in Figure \ref{map}. In fact, $\cloquence$ uses the Python grammar and parsing facilities directly. Several factors motivated this design decision. First, Python's syntax is widely credited as being particularly simple and readable, due to its use of significant whitespace and conventional mathematical notation. Python is one of the most widely-used languages in scientific computing, so its syntax is already familiar to much of the field. And significantly, a large ecosystem of tools already exist that work with Python files, such as code editors, syntax highlighters, style checkers and documentation generators. These can be used without modification to work with $\cloquence$ files. Therefore, by re-using an existing, widely-used grammar, we are able to satisfy many of the design criteria described in Section \ref{syntax} and the adoption criteria described in Section \ref{tools} without significant development effort.

\begin{figure}
\small{
\begin{verbatim}
import clq
from clq.backends.opencl import 
    OpenCL, get_global_id

@clq.fn(OpenCL)
def map(in, out, fn):
    gid = get_global_id(0)
    out[gid] = fn(in[gid])
\end{verbatim}}
\caption{Data-parallel higher-order map in $\cloquence$. The \texttt{get\_global\_id} function is an OpenCL primitive function.}
\label{map}

\end{figure}

\begin{figure}
\small{\begin{verbatim}
#pragma OPENCL EXTENSION cl_khr_fp64 : enable

__kernel void map_sin_dbl(__global double *in, 
                          __global double *out) 
{
    size_t gid = get_global_id(0);
    out[gid] = sin(in[gid]);
}
\end{verbatim}}
\caption{An OpenCL kernel function that maps the sin function over a vector of double-precision floating-point numbers in the global memory address space.}
\label{mapsinint}
\end{figure}

\subsection{Semantics}
$\cloquence$ is a compiled, statically-typed language. The extension used in Figure \ref{map} implements the full OpenCL type system, which includes features inherited from C99 like pointer arithmetic and OpenCL-specific constructs like vector types. In fact, compilation of Figure \ref{map} will produce code semantically equivalent, and with the same performance profile, as manually-written OpenCL code. Compilation produces OpenCL source code, so $\cloquence$, when used with this extension, inherits OpenCL's portability profile as well, as described in Section \ref{portability}.

In OpenCL itself, however, there is no way to write a function like \verb|map|. As described in Section \ref{hof}, OpenCL lacks support for higher-order functions or polymorphism (over the types of \verb|in| and \verb|out| in this case.) To map a function over an array, developers must create a separate function for every combination of argument types and for every transfer function. Figure 2 shows an OpenCL kernel that implements a map specifically for the \verb|sin| function applied to arrays of doubles in global memory. Note that because OpenCL does not support templates or function pointers, this is the most general solution possible without explicit code generation. However, even in CUDA, which does support templates, there is greater syntactic overhead for specializing a function with particular types than in $\cloquence$, where it happens implicitly.

To support this much more concise form for specifying kernels, $\cloquence$ uses techniques pioneered by functional programming languages. We briefly describe these below.

\subsubsection{Type, Kernel and Extension Inference}
As described in Section \ref{syntax}, professional end-users tend to prefer languages that do not require explicit type annotations for variables. These languages accomplish this by using dynamic typing, so that types are associated with values instead. Statically-typed functional programming languages have also eliminated type annotations in many cases, however, using a technique known as {\it type inference} \cite{tapl}.

Type inference is a technique for giving types to variables by examining how these variables are being used. As a simple example, in Figure \ref{map}, the variable \verb|gid| is being used to hold the result of calling the \verb|get_global_id| function so the type of \verb|gid| must be consistent with the return type of that function, \verb|size_t| (a device-dependent integer type.) Note, however, that in C99, many possible integer types can be given to \verb|gid| (in practice, a 32-bit integer is often used.) Picking a specific type may require further examining how \verb|gid| is used in the remainder of the function. 
To deal with this non-locality, type inference is typically cast as a constraint solving problem -- the type of each program variable is treated as an unknown and each use of the variable introduces a new constraint. A constraint-solving algorithm, generally a variant of unification, can be used to find a type assignment that satisfying all constraints.

$\cloquence$ uses a variant of this approach (modified to admit extensions to the type system, as we will discuss shortly) to conduct whole-function type inference. This approach differs from the HM algorithm, however, which conducts whole-program (or whole-file) inference and assigns a single type to function arguments. In $\cloquence$, types are {\it propagated} to functions from their calling sites (to enable structural polymorphism, described below) and inference occurs only within a function body.

In addition to type annotations, OpenCL introduces additional syntactic burdens. Developers must annotate functions that meet the requirements to be called from the host with the \verb|__kernel| attribute, and several types (notably, \verb|double|) and specialized functions require that an OpenCL extension be enabled with a \verb|#pragma| (as  can be seen in Figure \ref{mapsinint}.) The OpenCL extension we have developed automatically infers many of these annotations as well, consistent with the principle of conciseness described in Section \ref{syntax}.

\subsubsection{Structural Polymorphism}
In Section \ref{hof}, we discussed several strategies for achieving {\it polymorphism} -- the ability to create functions and data structures that operate over more than a single type. In $\cloquence$, all functions are implicitly polymorphic and can be called with arguments of {\it any type that supports the operations used by the function}. For example, in Figure \ref{map}, \verb|in| can be any type that supports indexing by a variable of type \verb|size_t| to produce a value of a type that can be passed into \verb|fn|, which must then produce a value consistent with indexing into \verb|out|. OpenCL pointer types are consistent with these constraints, for example. Although powerful, this also demonstrates a caveat of this approach -- that it is more difficult to give a function a concise signature, because arguments are constrained by capability, rather than to a single type \cite{malayeri2009structural}.

Structural typing can be compared to the approach taken by dynamically-typed languages that rely on ``duck typing''. It is more flexible than the parametric polymorphism found in many functional languages and in languages like Java (which only allow polymorphic functions that are valid for {\it all} possible types), but is of comparable strength to the template system found in C++. It can be helpful to think of each function as being preceded by an implicit template header that assigns each argument its own unique type parameter. At function call sites, these parameters are implicitly specialized with the types of the provided arguments. This choice is again motivated by the criteria of conciseness given in Section \ref{syntax}.

\subsubsection{Higher-Order Functions}
Section \ref{hof} also discusses the value of higher-order functions. The OpenCL extension supports higher-order functions, despite the absence of function pointers from OpenCL itself, by explicitly passing functions as constants at compile-time. In other words, each function {\it uniquely inhabits} a type corresponding to it.

Functions are not, however, {\it first-class} -- references to functions cannot be stored in memory, for example. Most use cases that we have considered thus far have never needed to do so, however. Compared to the function pointer approach used in other C-based languages, this approach is more efficient. It can, however, result in ``code bloat'', since a new copy of the higher-order function is created for each unique function argument passed to it. First-class functions can theoretically be simulated using an auxiliary lookup function in OpenCL, and we leave clean support for this in $\cloquence$ as open future work.

\subsection{Active Libraries in $\cloquence$}
In addition to adopting Python's grammar, $\cloquence$ uses the Python language itself as a {\it compile-time metalanguage}. Every $\cloquence$ file is, at the top-level, a Python script. This script is executed during compilation and has access to a number of powerful capabilities that allow developers to influence the compilation process and manipulate programs directly. Libraries that have such capabilities has been called {\it active libraries} in prior proposals \cite{activelibraries}. A number of  projects, such as Blitz++, have taken advantage of the C++ preprocessor and template-based metaprogramming system to implement domain-specific optimizations. In $\cloquence$, we replace these brittle mini-languages with a general-purpose language. This allows for several interesting uses that we discuss in the following sections.

\subsection{Module System}
In Figure \ref{map}, the only compile-time operations are library imports and a declaration of a $\cloquence$ function. However, this already demonstrates an important benefit of this approach: $\cloquence$ uses Python's simple but flexible module system. As a consequence, libraries written using $\cloquence$ can directly utilize the module distribution infrastructure available for Python, partially relieving the issues discussed in Section \ref{modularity}.

\subsection{Compilation and Invocation}
$\cloquence$ functions can be invoked in one of two ways: from any host language supporting OpenCL by using the standalone compiler, or directly from Python itself. 
\begin{figure}\small{\begin{verbatim}
from figure1 import map
from clq.backends.opencl import double, sin

map_sin_double = map.specialize(
    double.global_ptr, double.global_ptr, 
    sin.clq_type)
\end{verbatim}}
\caption{Programmatically specializing the \texttt{map} function from Figure \ref{map} for use by the standalone compiler to produce code equivalent to Figure \ref{mapsinint}.}
\label{specialization}
\end{figure}

\subsubsection{Programmatic Specialization and Compilation}
The standalone $\cloquence$ compiler, \verb|clqcc|, generates source code from $\cloquence$ source files. To do so, however, users must {\it specialize} any {\it externally callable} functions with their specific types. Figure \ref{specialization} shows how the specific OpenCL function given in Figure \ref{mapsinint} can be recovered from the generic specification of \verb|map| given in Figure \ref{map}. The \verb|specialize| method of a generic function like \verb|map| creates a new function that is concrete -- that is, with concrete argument types.

The standalone compiler operates by first executing the module, then iterating over the externally available variables in the module's environment to extract the OpenCL functions and any other functions and constructs that they depend on. This allows developers who wish to use the host language of their choice, rather than Python, to continue writing library functions generically, in one place, only writing small stub modules to specialize them as needed.

\subsubsection{Direct Invocation from Python}\label{direct}
As discussed in the introduction, a common usage scenario in scientific computing involves the use of a dynamic, high-level language for workflow orchestration and exploratory visualization and analysis, paired with a low-level language for performance-critical sections. Although the extension mechanism we describe shortly will permit the development of a broad array high-level language features in the future, current extensions, such as the OpenCL extension we have been describing, have been designed to make low-level programming simpler. For this reason, the OpenCL extension includes support for directly invoking $\cloquence$ functions from Python scripts with minimal overhead.

This feature builds on top of an established library, \verb|pyopencl|, that wraps the OpenCL host API and integrates it with the popular \verb|numpy| numerics library in Python. Buffers (arrays) in \verb|pyopencl| do not retain type information, so the $\cloquence$ extension adds type information, mirroring the API of \verb|numpy| arrays. It also hides many of the most verbose components of the OpenCL interface by providing reasonable defaults, such as an implicit global context and a default queue associated with each context.

Using these simplifications, generic $\cloquence$ OpenCL functions like \verb|map| can be called directly, without specialization as in the previous section. Figure \ref{py} shows a full OpenCL program written using these bindings. By way of comparison, the same program written using OpenCL directly is two orders of magnitude larger and correspondingly more complex. Not shown are several additional conveniences, such as delegated kernel sizing and In and Out constructs that can reduce the size and improve the clarity of this code further -- due to space constraints, the reader is referred to the language documentation for additional details.

\begin{figure}
\small{\begin{verbatim}
from figure1 import map
import numpy as np
import clq.opencl.pyopencl as cl

ctx = cl.Context.for_device(0, 0)
d_in = ctx.to_device(np.ones(1024))
d_out = ctx.alloc(like=d_x)
map(d_in, d_out, clq.opencl.sin, 
    global_size=d_in.shape, local_size=(128,))
out = ctx.from_device(d_out)
\end{verbatim}}
\caption{A full OpenCL program using the $\cloquence$ Python bindings, including data transfer and kernel invocation.}
\label{py}
\end{figure}

\subsection{Code Generation in $\cloquence$}
Metaprogramming refers to the practice of writing programs that manipulate other programs. There are a number of use cases for this technique, including domain-specific optimizations and code generation for programs with a repetitive structure that cannot easily be packaged using available abstractions. OpenCL in particular relies on code generation as a fundamental mechanism, partially justifying the lack of support for higher-order programming.

$\cloquence$ supports code generation from directly within the metalanguage. A cl.oquence function can be constructed from a string containing its source using the \verb|clq.from_source| function. It can also be constructed directly from a Python abstract syntax tree, available via the standard \verb|ast| package, using the \verb|clq.from_ast| function. We discuss a use case for code generation for simulation orchestration in $\cloquence$ in Section 4.

\subsection{Active Language Extensions}
We now turn our attention to the extension mechanism used by $\cloquence$, as motivated by section \ref{extensions}. The $\cloquence$ type checker and compiler interacts directly with active libraries over an interface that gives substantial control over type checking and translation of an operation to a {\it type definition} associated with one of its operands according to a fixed dispatch protocol.

\subsubsection{Dispatch Protocol}
\begin{figure}\small{
\begin{verbatim}
class PtrType(Type):
  def __init__(self, target_type, addr_space):
    self.target_type = target_type
    self.addr_space = addr_space

  def verify_Subscript(self, context, node):
    slice_type = context.validate(node.slice)
    if isinstance(slice_type, IntegerType):
      return self.target_type
    else: 
      raise TypeError('<error message here>')

  def translate_Subscript(self, context, node):
    value = context.translate(node.value)
    slice = context.translate(node.slice)
    return copy_node(node, 
      value = value,
      slice = slice,    
      code = value.code + '['+slice.code+']')
\end{verbatim}}
\caption{A portion of the implementation of OpenCL pointer types implementing subscripting logic using the $\cloquence$ extension mechanism.}
\label{pointers}
\end{figure}

When the compiler encounters an expression, it must first verify that it type checks, then generate code for it. Rather than containing fixed logic for this, however, the compiler defers this responsibility to the {\it type} of a subexpression whenever possible. Below are example from the $\cloquence$ dispatch protocol. Due to space constraints, we do not list the entire dispatch protocol.
\begin{itemize}
\item Responsibility over a {\bf unary operation} like \verb|-x| is handed to the type assigned to the operand, \verb|x|.
\item Responsibility over {\bf binary operations} is first handed to the type assigned to the left operand. If it indicates that it does not understand the operation, the type assigned to the right operand is handed responsibility, with a different method call\footnote{Note that this operates similarly to Python's operator overloading protocol.}.
\item Responsibility over {\bf attribute access}, \texttt{obj.attr}, and {\bf subscript access}, \texttt{obj[idx]}, is handed to the type assigned to \texttt{obj}.
\end{itemize}

\subsubsection{Verification and Translation}
A type in $\cloquence$ is a metalanguage {\it instance} of a Python class deriving from \verb|clq.Type|. The compiler hands control over an expression to a type by calling a method corresponding to the syntactic form of the expression according to the above dispatch protocol. Figure \ref{pointers} gives the verification and translation logic governing subscript access for pointer types in OpenCL.

During the verification phase, the type of the primary operand, as determined by the dispatch protocol, is responsible for assigning a type to the expression as a whole. In this case, it recursively assigns a type to the slice operand. If it is assigned an integer type, the pointer's target type is assigned to the expression as a whole. Otherwise, a type error is raised with the provided error message (an ability that can be used by extension authors to satisfy the error message criteria described in Section \ref{errors}.)

If verification succeeds, the compiler must subsequently translate the $\cloquence$ expression into an expression in the output language, here OpenCL. It does so by again applying the dispatch protocol to call a method prefixed with \verb|translate_|. This method is responsible for returning a copy of the expression's ast node with an additional attribute, \verb|code|, containing the source code of the translation. In this case, it is simply a direct translation to the corresponding OpenCL attribute access, using the recursively-determined translations of the operands. More sophisticated abstractions may insert arbitrarily complex statements and expressions. The context also provides some support for non-local effects, such as new top-level declarations (not shown.)

\subsubsection{Modular Backends}\label{backends}
Thus far, we have discussed using OpenCL as a backend with $\cloquence$. The OpenCL extension is the most mature as of this writing. However, $\cloquence$ supports the introduction of new backends in a manner similar to the introduction of new types, by extending the \verb|clq.Backend| base class. Backends are provided as the first argument to the \verb|@clq.fn| decorator, as can be seen in Figure \ref{map}. 
Backends are responsible for some aspects of the grammar that do not admit simple dispatch to the type of a subterm, such as number and string literals or basic statements like \verb|while|.

In addition to the OpenCL backend, preliminary C99 and CUDA backends are available (with the caveat that they have not been as fully developed or tested as of this writing.) Backends not based on the C family are also possible, but we leave such developments for future work.

\subsubsection{Use Cases}
The development of the full OpenCL language using only the extension mechanisms described above provides evidence of the power of this approach. Nothing about the core language was designed specifically for OpenCL. However, to be truly useful, as described in Sections \ref{multiparadigm} and  \ref{extensions}, the language must be able to support a wide array of primitive abstractions. We briefly describe a number of other abstractions that may be possible using this mechanism. Many of these are currently available either via inconvenient libraries or in standalone languages. With the $\cloquence$ extension mechanism, we hope to achieve robust, natural implementations of many of these mechanisms within the same language.

\paragraph{Partitioned Global Address Spaces}
A number of recent languages in high-performance computing have been centered around a partitioned global address space model, including UPC, Chapel, X10 and others. These languages provide first-class support for accessing data transparently across a massively parallel cluster, which is verbose and poorly supported by standard C. The extension mechanism of $\cloquence$ allows inelegant library-based approaches such as the Global Arrays library to be hidden behind natural wrappers that can use compile-time information to optimize performance and verify correctness. We have developed a prototype of this approach using the C backend and hope to expand upon it in future work.

\paragraph{Other Parallel Abstractions}
A number of other parallel abstractions, some of which are listed in \ref{multiparadigm}, also suffer from inelegant C-based implementations that spurred the creation of standalone languages. A study comparing a language-based concurrency solution for Java with an equivalent, though less clean, library-based solution found that language support is preferable but leads to many of the issues we have described \cite{cave2010comparing}. The extension mechanism is designed to enable library-based solutions that operate as first-class language-based solutions, barring the need for particularly exotic syntactic extensions.

\paragraph{Domain-Specific Type Systems}
$\cloquence$ is a statically-typed language, so a number of domain-specific abstractions that promise to improve verifiability using types, as discussed in Section \ref{verifiability}, can be implemented using the extension mechanism. We hope that this will allow advances from the functional programming community to make their way into the professional end-user community more quickly, particularly those focused on scientific domains (e.g. \cite{conf/cefp/Kennedy09}).

\paragraph{Specialized Optimizations}
In many cases, code optimization requires domain-specific knowledge or sophisticated, parametrizable heuristics. Existing compilers make implementing and distribution such optimizations difficult. With active libraries in $\cloquence$, optimizations can be distributed directly with the libraries that they work with. For instance, we have implemented substantial portions of the NVidia GPU-specific optimizations described in \cite{yang2010gpgpu} as a library that uses the extension mechanism to track affine transformations of the thread index used to access arrays, in order to construct a summary of the memory access patterns of the kernel, which can be used both for single-kernel optimization (as in \cite{yang2010gpgpu}) and for future research on cross-kernel fusion and other optimizations.

\paragraph{Instrumentation}
Several sophisticated feedback-directed optimizations and adaptive run-time protocols require instrumenting code in other ways. The extension mechanism enables granular instrumentation based on the form of an operation as well as its constituent types, easing the implementation of such tools. This ability could also be used to collect data useful for more rigorous usability and usage studies of languages and abstractions, and we plan on following up on this line of research going forward.

\subsection{Availability}
$\cloquence$ is available under the LGPL license and is developed openly and collaboratively using the popular Github platform\footnote{https://github.com/cyrus-/cl.oquence}. Documentation and other learning material is being made available at http://cl.oquence.org/.

\section{Case Study: Neurobiological Circuit Simulation}
An important criteria that practitioners use to evaluate a language or abstraction, as discussed in Section \ref{social}, is whether significant case studies have been conducted with it. In this section, we briefly (due to space limitations) discuss an application of the $\cloquence$ OpenCL library, Python host bindings and code generation features for developing a modular, high-performance scientific simulation library used to simulate  thousands of parallel realizations of a spiking neurobiological circuit on a GPU.

\subsection{Background}
A neural circuit can be modeled as a network of coupled differential equations, where each node corresponds to a single neuron. Each neuron is modeled using one or more ordinary differential equations. These equations capture the dynamics of physically important quantities like the cell's membrane potential or the conductance across various kinds of ion channels and can take many forms \cite{neurobook}. Single simulations can contain from hundreds to tens of millions of neurons each, depending on the specific problem being studied. In some cases, such as when studying the effects of noise on network dynamics or to sweep a parameter space, hundreds or thousands of realizations must be generated. In these cases, care must be taken to only probe the simulation for relevant data and process portions of it as the simulation progresses, because the amount of data generated is often too large to store in its entirety for later analysis.

The research group we discuss here (of which the first author was a member) was studying a problem that required running up to 1,000 realizations of a network of between 4,000 and 10,000 neurons each. An initial solution to this problem used the Brian framework, written in Python, to conduct these simulations on a CPU. Brian was selected because it allowed the structure of the  simulation to be specified in modular and straightforward manner. This solution required between 60 and 70 minutes to conduct the simulations and up to 8 hours to analyze the data each time a parameter of the simulation was modified.

Unsatisfied with the performance of this approach, the group developed an accelerated variant of the simulation using C++ and CUDA. Although this produced significant speedups, reducing the time for a simulation by a factor of 40 and the runtime of the slowest analyses by a factor of 200, the overall workflow was also significantly disrupted. In order to support the many variants of models, parameter sets, and probing protocols, C preprocessor flags were necessary to selectively include or exclude code snippets. This quickly led to an incomprehensible and difficult to maintain file structure. Moreover, much of the simpler data analysis and visualization was conducted using Python, so marshalling the relevant data between processes also became an issue.

\subsection{The {\sf cl.egans} Simulation Library}
In order to eliminate these issues while retaining the performance profile of the GPU-accelerated code, the project was ported to $\cloquence$. Rather than using preprocessor directives to control the code contained in the final GPU kernels used to execute the simulation and data analyses, the group was able to develop a more modular  library called {\sf cl.egans}\footnote{...after {\it c. elegans}, a model organism in neuroscience} based on the language's compile-time code generation mechanism and Python and OpenCL bindings.

{\sf cl.egans} leverages Python's object-oriented features to enable modular, hierarchical simulation specifications. For example, Figure \ref{spec} shows an example where a neuron model (\verb|ReducedLIF|) is added to the root of the simulation, a synapse model (\verb|ExponentialSynapse|) is then added to it, and its conductance is probed in the same way, by adding a probe model as a child of the synapse model. If interleaved analysis needed to be conducted as well, it would be specified in the same way.

Implementations of these classes do not evaluate the simulation logic directly, but rather contain methods that generate $\cloquence$ source code for insertion at various points, called {\it hooks}, in the final simulation kernel. The hook that code is inserted into is determined by the method name, and code can be inserted into any hook defined anywhere upstream in the simulation tree. New hooks can also be defined in these methods and these become available for use by child nodes. Figure \ref{impl} shows an example of a class that inserts code in the \verb|model_code| hook and defines several new hooks. This protocol is closely related to the notion of {\it frame-oriented programming}. Although highly modular, this strategy avoids the performance penalties associated with standard object-oriented methodologies via code generation.

Compared to a similar protocol targeting OpenCL directly, the required code generation logic is significantly simpler because it enables classes like \verb|StateVariable| to be written generically for all types of state variables, without carrying extra parameters and {\it ad hoc} logic to extract and compute the result types of generated expressions. Moreover, because types are first-class objects in the metalanguage, they can be examined during the memory allocation step to enable features like fully-automatic parallelization of multiple realizations across one or more devices, a major feature of {\sf cl.egans} that competing frameworks cannot easily offer.

Once the kernel has been generated and memory has been allocated, the simulation can be executed directly from Python using the bindings described in Section \ref{direct}. The results of this simulation are immediately available to the Python code following the simulation and can be visualized and further analyzed using standard tools. Once the computations are complete, the Python garbage collector is able to handle deallocation of GPU memory automatically (a feature of the underlying \verb|pyopencl| library \cite{klockner2011pycuda}.)

Using this $\cloquence$-based framework, the benefits of the Brian-based workflow were recovered without the  corresponding decrease in performance relative to the previous CUDA-based solution, leading ultimately to a satisfying solution for the group conducting this research.

\begin{figure}
\small{
\begin{verbatim}
# Create the root node of the simulation
sim = Simulation(ctx, n_timesteps=10000)
neurons = ReducedLIF(sim, count=N, tau=20.0)
e_synapse = ExponentialSynapse(neurons, 'ge',
    tau=5.0, reversal=60.0)
probe = StateVariableProbeCopyback(e_synapse)
\end{verbatim}}
\caption{An example of a nested simulation tree, showing that specifying a simulation is both simple and modular.}
\label{spec}
\end{figure}

\begin{figure}\small{
\begin{verbatim}
class SpikingModel(Model):
  """Base class for spiking neuron models."""
  def in_model_code(self, g):
    """
    idx_state = idx_model + (realization_num - 
        realization_start)*count
    """ << g  # << appends to code generator, g
    self.insert_cg_hook("read_incoming", g)
    self.insert_cg_hook("read_state", g)
    self.insert_cg_hook("calculate_inputs", g)
    self.insert_cg_hook("state_calculations", g)
    self.insert_cg_hook("spike_processing", g)
  # ...
\end{verbatim}}
\caption{An example of a hook that inserts code and also inserts new, nested hooks for downstream simulation nodes  below that.}
\label{impl}
\end{figure}

\section{Conclusion}
Professional end-users demand much from new languages and abstractions. In this paper, we began by generating a concrete, detailed set of design and adoption criteria that we hope will be of broad interest and utility to the research community. Based on these constraints, we designed a new language, $\cloquence$, making several pragmatic design decisions and utilizing advanced techniques, including type inference, structural typing, compile-time metaprogramming and active libraries, to uniquely satisfy many of the criteria we discuss, particularly those related to extensibility. We validated the extension mechanism with a mature implementation of  the entirety of the OpenCL type system, as well as preliminary implementations of some other features. Finally, we demonstrated that this language was useful in practice, drastically improving performance without negatively impacting the high-level scientific workflow of a large-scale neurobiological circuit simulation project. Going forward, we hope that $\cloquence$ (or simply the key techniques it proposes, by some other vehicle) will be developed further by the community to strengthen the foundations upon which new abstractions are implemented and deployed into professional end-user development communities.

% An example of a floating figure using the graphicx package.
% Note that \label must occur AFTER (or within) \caption.
% For figures, \caption should occur after the \includegraphics.
% Note that IEEEtran v1.7 and later has special internal code that
% is designed to preserve the operation of \label within \caption
% even when the captionsoff option is in effect. However, because
% of issues like this, it may be the safest practice to put all your
% \label just after \caption rather than within \caption{}.
%
% Reminder: the "draftcls" or "draftclsnofoot", not "draft", class
% option should be used if it is desired that the figures are to be
% displayed while in draft mode.
%
%\begin{figure}[!t]
%\centering
%\includegraphics[width=2.5in]{myfigure}
% where an .eps filename suffix will be assumed under latex, 
% and a .pdf suffix will be assumed for pdflatex; or what has been declared
% via \DeclareGraphicsExtensions.
%\caption{Simulation Results}
%\label{fig_sim}
%\end{figure}

% Note that IEEE typically puts floats only at the top, even when this
% results in a large percentage of a column being occupied by floats.

% An example of a double column floating figure using two subfigures.
% (The subfig.sty package must be loaded for this to work.)
% The subfigure \label commands are set within each subfloat command, the
% \label for the overall figure must come after \caption.
% \hfil must be used as a separator to get equal spacing.
% The subfigure.sty package works much the same way, except \subfigure is
% used instead of \subfloat.
%
%\begin{figure*}[!t]
%\centerline{\subfloat[Case I]\includegraphics[width=2.5in]{subfigcase1}%
%\label{fig_first_case}}
%\hfil
%\subfloat[Case II]{\includegraphics[width=2.5in]{subfigcase2}%
%\label{fig_second_case}}}
%\caption{Simulation results}
%\label{fig_sim}
%\end{figure*}
%
% Note that often IEEE papers with subfigures do not employ subfigure
% captions (using the optional argument to \subfloat), but instead will
% reference/describe all of them (a), (b), etc., within the main caption.

% An example of a floating table. Note that, for IEEE style tables, the 
% \caption command should come BEFORE the table. Table text will default to
% \footnotesize as IEEE normally uses this smaller font for tables.
% The \label must come after \caption as always.
%
%\begin{table}[!t]
%% increase table row spacing, adjust to taste
%\renewcommand{\arraystretch}{1.3}
% if using array.sty, it might be a good idea to tweak the value of
% \extrarowheight as needed to properly center the text within the cells
%\caption{An Example of a Table}
%\label{table_example}
%\centering
%% Some packages, such as MDW tools, offer better commands for making tables
%% than the plain LaTeX2e tabular which is used here.
%\begin{tabular}{|c||c|}
%\hline
%One & Two\\
%\hline
%Three & Four\\
%\hline
%\end{tabular}
%\end{table}

% Note that IEEE does not put floats in the very first column - or typically
% anywhere on the first page for that matter. Also, in-text middle ("here")
% positioning is not used. Most IEEE journals/conferences use top floats
% exclusively. Note that, LaTeX2e, unlike IEEE journals/conferences, places
% footnotes above bottom floats. This can be corrected via the \fnbelowfloat
% command of the stfloats package.

% conference papers do not normally have an appendix

% use section* for acknowledgement
% trigger a \newpage just before the given reference
% number - used to balance the columns on the last page
% adjust value as needed - may need to be readjusted if
% the document is modified later
%\IEEEtriggeratref{8}
% The "triggered" command can be changed if desired:
%\IEEEtriggercmd{\enlargethispage{-5in}}

% references section

% can use a bibliography generated by BibTeX as a .bbl file
% BibTeX documentation can be easily obtained at:
% http://www.ctan.org/tex-archive/biblio/bibtex/contrib/doc/
% The IEEEtran BibTeX style support page is at:
% http://www.michaelshell.org/tex/ieeetran/bibtex/
%\bibliographystyle{IEEEtran}
% argument is your BibTeX string definitions and bibliography database(s)
%\bibliography{IEEEabrv,../bib/paper}
%
% <OR> manually copy in the resultant .bbl file
% set second argument of \begin to the number of references
% (used to reserve space for the reference number labels box)
\bibliographystyle{IEEEtran}
\bibliography{research}
%
%\begin{thebibliography}{1}
%
%\bibitem{IEEEhowto:kopka}
%H.~Kopka and P.~W. Daly, \emph{A Guide to \LaTeX}, 3rd~ed.\hskip 1em plus
%  0.5em minus 0.4em\relax Harlow, England: Addison-Wesley, 1999.
%
%\end{thebibliography}

% that's all folks
\end{document}

### 2012-09-04 ipdps13/clq-ipdps13.tex
Abstract (may be too long?)

@@
 %
-%\usepackage{url}
+\usepackage{url}
 % url.sty was written by Donald Arseneau. It provides better support for
@@
 
-\author{\IEEEauthorblockN{Cyrus Omar, Jonathan Aldrich}
+\author{\IEEEauthorblockN{Cyrus Omar, Nathan Fulton, Sharschchandra Bidargaddi and Jonathan Aldrich}
 \IEEEauthorblockA{School of Computer Science\\
@@
 Pittsburgh, PA, USA\\
-\texttt{\{comar,jonathan.aldrich\}@cs.cmu.edu}}}
+\url{http://www.acelang.org/}
+}}
 
@@
 \begin{abstract}
-In this paper, we introduce the Ace programming language and demonstrate its practicality as a foundational tool for both high-performance computing researchers and practitioners.
-Ace is a statically-typed language that shares a syntax with Python and uses the Python language itself as a multi-purpose compile-time metalanguage.
-Ace has a minimal core that delegates control over many aspects of compilation to user modules via a novel extension mechanism called {\em active type-checking and translation (\ATT)}.
-Specifically, users specify new primitive types and operations by equipping type definitions -- first-class objects in the metalanguage -- with compile-time methods.
-The compiler invokes these methods when type checking and translating expressions, according to a fixed type dispatch protocol associated with each syntactic form.
+In this paper, we introduce the Ace compilation environment and demonstrate its practicality as a foundational tool in both high-performance computing research and in practice.
+Ace consists of a statically-typed language that shares syntax with Python and uses the Python language itself as a multi-purpose compile-time metalanguage.
+Unlike standard languages, however, the Ace compiler consists only of a minimal core that delegates control over most aspects of compilation to user libraries using a novel extension mechanism that we call {\em active type-checking and translation (\ATT)}.
+In particular, users can specify new primitive types and operations by equipping type definitions, which are first-class objects in the metalanguage, with special methods.
+The compiler then invokes these methods when type checking and translating expressions, according to a fixed type dispatch protocol associated with each syntactic form.
+Translation targets a user-defined backend language, and the safety of user-defined translations can be checked automatically.
+Using this mechanism, we implement the primitives of the C99 and OpenCL languages as simple modules within Ace.
+We go on show examples of other interesting primitive parallel  abstractions that build on top of these basic modules, demonstrating the foundational role that an actively-typed system like Ace can play in support of research on language-based parallel programming abstractions.
+
+In addition to \ATT, Ace supports more conventional forms of structured and unstructured metaprogramming  and integrates directly with the numeric capabilities of Python itself.
+We use these features to design a scientific simulation framework that allows users to modularly specify and orchestrate the execution of parameterized families of simulations on clusters of GPUs.
+This framework is used to successfully conduct large-scale, high-performance neuroscience simulations, providing  evidence that Ace can be used in practice today.
 
@@
 %programming languages, heterogeneous computing, metaprogramming, code generation
-
 %\end{IEEEkeywords}
@@
 Finally, conclude in Section 6 with future directions for both the $\cloquence$ language and the research community.
-
-\section{Design and Adoption Criteria}
-The primary purpose of this paper is to describe the design of a programming language, $\cloquence$.
-Unfortunately, as in many areas of design, it can be difficult to objectively evaluate the merit of such efforts.
-To better support such evaluations, researchers in design disciplines typically develop a set of high-level  design criteria that serve as a rubric for evaluating and guiding their design efforts.
-In programming language design, particularly for scientific and high-performance computing, there have been few concerted efforts to develop a coherent set of design criteria that capture the needs of the targeted developer communities.
-Similarly, there have been few treatments of adoption criteria for new languages and abstractions in practice, an important issue given the slow rates of adoption today.
-We have purposefully separated these criteria from the specific language we describe in Section 3 in the hopes of starting a productive discussion within the community.
-
-\subsection{Professional End-User Developers}
-An important first step is to identify the developer communities targeted by our design efforts.
-Researchers in software engineering typically distinguish between {\it end-user developers} and {\it professional developers}.
-End-user developers have little formal training relevant to the tools that they use, and the tools themselves are relatively simple (spreadsheets, for example).
-Professional developers, on the other hand, have explicit training or extensive experience with software engineering and software development techniques.
-Scientists and engineers do not cleanly fall into either of these groups, however -- although formal training in software development is rare \cite{oai:open.ac.uk.OAI2:17673}, members of these groups are more technically literate and demanding than typical end-users.
-For this reason, researchers have proposed a third group, {\it professional end-user developers}, to describe ``people working in highly technical, knowledge rich professions, such as financial mathematicians, scientists and engineers, who develop their own software in order to advance their own professional goals.
-\cite{segal2007some}''
-As such, we consider the literature on novice and end-user programming (cf.
-\cite{pane1996usability}) in addition to that on programming language design for professional developers to justify the criteria we develop below.
-Language usability and adoption are not commonly studied topics, so many of the hypotheses we present below also rely on informal observations about languages and tools that have been successful in the past.
-
-\subsection{Design Criteria}
-We define design criteria as aspects of a language's design, rather than its implementation, that help developers express their intent naturally and correctly.
-
-\subsubsection{Concise, Familiar and Readable Syntax}\label{syntax}
-Despite a considerable amount of evidence pointing toward the value of a well-designed syntax, particularly for end-users in specialized domains \cite{pane1996usability}, the issue is sometimes marginalized.
-It is likely the case, however, that some libraries and languages experience low adoption due in  part due to the use of a verbose, unfamiliar or unreadable syntactic style.
-
-A simple and concise syntax is common to most of the high-level languages that are widely used in scientific computing.
-Cordy identifies the principle of conciseness with elimination of redundancy and the availability of reasonable defaults \cite{cordy1992hints}.
-The high-level languages listed in the introduction have all either made optional, or removed entirely, much of the syntactic overhead characteristic of low-level languages, such as explicit variable declarations and extensive headers or preambles that contain information that can be inferred from the body of the program in most cases.
-They also typically feature simple array indexing syntax and concise literal forms for common data structures like arrays, sets and maps.
-A concise and minimal syntax eliminates unnecessary keystrokes and keeps more code visible on screen at a time.
-Studies have shown that there may be a correlation between lines of code entered and overall error rate, independent of other factors \cite{el2001confounding}.
-
-However, one can also go too far, so these benefits must be balanced with concerns about readability and familiarity.
-A number of principles have been proposed to operationalize the notion of code readability.
-The most widely-used collection of principles are Green's cognitive dimensions \cite{green1996usability}.
-Of particular relevance is the notion of self-consistency, which serves to ensure that similar forms have similar meaning and that there are few subtle or context-dependent distinctions that users must be mindful of.
-Another important principle has been called {\it closeness of mapping}, expressing the value of a close correspondence between mental models and the formal model as expressed concretely using the language.
-
-Most researchers become familiar with formal notation by studying mathematics.
-In nearly all commonly used languages in scientific computing, the common mathematical notation is used whenever possible.
-In contrast, many academic languages have settled on alternative notational styles.
-For example, the LISP family of languages uses a highly uniform list-based notation, while most functional languages typically borrow notation for function invocation from the lambda calculus, using the form \verb|f x| rather than \verb|f(x)| for function invocation.
-Although both of these styles have certain benefits, they can impose mental burdens on users who continue to mentally translate them into more familiar notation \cite{anderson1985novice}.
-Although these and related difficulties can decrease with experience, they remain an important barrier for new users to these languages.
-
-Syntactic cues like whitespace and typography that do not have a formal meaning but rather exist to assist developers are called secondary notation \cite{green1990programming}.
-Most languages are whitespace-insensitive, while others (notably, Python) enforce consistent uses of whitespace, both in pursuit of consistency and as a technique to eliminate the need for block delimiters.
-Additionally, a few languages, such as Mathematica, have support for more typographically-rich mathematical notation via a structural editing interface.
-These techniques appear to be helpful, although we do not know of formal studies that provide evidence for this claim.
-
-\subsubsection{Support for Multiple Paradigms and Abstractions}\label{multiparadigm}
-Although the difficulties of low-level and parallel programming are widely acknowledged, there is little consensus on which high-level abstractions are most appropriate for easing this burden.
-Indeed, it appears likely that no one abstraction will emerge triumphant over the others.
-Although library-based abstractions are useful, they often suffer from limitations of the language, particularly if they require compile-time support.
-Primitive language support for a parallel abstraction has been shown to be more usable than a library-based implementation in at least one case \cite{cave2010comparing}.
-Languages that have been designed specifically to explore a single abstraction as a core language feature often see more limited adoption than library-based implementations because they remain difficult to use in circumstances for which an alternative abstraction is more appropriate.
-Examples of abstractions where language-based implementations remain less frequently used than library-based implementations include the actor model (e.g.
-Erlang), software transactional memory \cite{harris2003language}, and global arrays (e.g.
-UPC).
-As such, it is important that a language support several modes of operation, ideally without excessively favoring one over another.
-
-\subsubsection{Extensibility}\label{extensions}
-
-Although support for multiple paradigms and primitive abstractions can be built into a language design, this leaves control in the hands of the language designer.
-Novel or domain-specific constructs that may be useful to a small number of users or in rare situations can be difficult to develop and distribute for this reason, leading to the proliferation of new languages as  described above.
-Language extensibility mechanisms support these use cases by giving users the ability to develop new abstractions and constructs that behave as if they were primitive constructs.
-If a mechanism is powerful enough, nearly all language constructs may be implemented using it, greatly simplifying the core semantics of a language.
-
-Dynamic languages commonly rely on mechanisms like operator overloading and metaobject protocols \cite{Kiczales91} to provide extensibility via indirection.
-For example, the Python language allows objects to overload nearly every operator and as well as operations like attribute lookup (\verb|obj.attr|) and assignment.
-This mechanism is used by a number of libraries to create a more natural interface to a low-level API (e.g.
-\verb|pycuda|).
-More open-ended dynamic mechanisms, such as programmatic macros, have also been widely studied but as of yet have seen little adoption in languages used by professional end-users.
-
-Language extensibility for statically-typed languages, on the other hand, remains an active research area.
-Compile-time metaprogramming systems, which are related to programmatic macros, have been developed for a number of languages (e.g.
-Template Haskell \cite{sheard2002template}) but these too have seen relatively limited development or adoption thus far.
-In Section 3, we introduce a novel static language extension mechanism that uses type-based dispatch rather than direct manipulation of syntax trees.
-
-Some researchers have advocated the use of compiler extension mechanisms, rather than mechanisms built into a language itself (cf.
-\cite{clements2008comparison}).
-Modern compilers now offer some support for front-end language extensions, although often these  can be quite difficult to use.
-Although potentially powerful, this approach can also lead to issues when extensions are not easily composable or when multiple compilers exist for a language.
-Back-end extensions (that is, extensions that preserve the semantics of the language, improving only performance) are better supported in modern compilers.
-
-Domain-specific language frameworks are a related approach that can serve many of the same goals as language-based  extension mechanisms \cite{fowler2010domain}.
-These tools ease the development of ``little languages'' and allow for the development and distribution of language features as modules.
-Domain experts use these tools to develop highly specialized languages that capture domain requirements precisely and allow for natural and concise specifications of scientific models and other structures.
-A major issue with this approach arises at language boundaries -- interoperability between different domain-specific languages is difficult due to feature mismatches.
-Another issue is that domain-specific languages often outgrow their initial implementations and begin to need increasingly powerful general-purpose features.
-
-All extensibility mechanisms must be used carefully.
-Indeed, inexperienced developers can abuse mechanisms like operator overloading to create inscrutable interfaces that cause more problems than they solve.
-Some languages (notably Java) have taken up the philosophy that even simple language extension mechanisms should not be in the hands of end-users for this reason.
-Although this continues to be debated, we argue that extensibility is crucial for parallel and scientific programming in particular due to the significant levels of ongoing research into new parallel abstractions and the diverse set of other domain-specific use cases.
-
-\subsubsection{Support for Higher-Order Constructs}\label{hof}
-A number of parallel programming data structures and algorithms are of higher-order, meaning that they take other functions or types as arguments or parameters.
-We argue that support for higher-order programming is critical to language usability in our target domains.
-
-Well-known examples of functions that operate using other functions include fundamental parallel primitives like map, reduce and scan.
-Fortunately, most modern languages now support using functions as values.
-Languages like C and Fortran implement this using function pointers, although at considerable syntactic expense.
-However, unlike functional languages and most dynamic languages, they do not allow anonymous functions (that is, functions that are defined as inline expressions rather than statements), nor functions that close over variables in the surrounding scope.
-The most recent revision of C++ now has support for closures and closures are also being considered for a future revision of Java.
-OpenCL does not come with any support for higher-order functions, an issue we explore further in Section 3.
-
-Functions and types that are parameterized by types are often referred to as {\it polymorphic} or {\it generic}.
-C++ supports this using its template system.
-Java, ML and Haskell support a simpler form of {\it parametric polymorphism}, and Haskell also supports a more flexible {\it type class} system.
-C++, Java and Fortran also support {\it function overloading}, allowing multiple versions of a function that differ only according to their argument types.
-C and OpenCL do not support any form of polymorphism or user-defined function overloading.
-
-Dynamic languages do not require that variables be assigned a type, so generic functions are written using run-time checks that ensure that a particular value supports the specific interface that a function expects.
-This is sometimes known as ``duck typing'' and is generally considered as a useful feature, due to its considerable flexibility.
-A promising approach that resembles duck typing, while operating statically, is known as {\it structural typing} \cite{malayeri2009structural}.
-
-\subsubsection{Modularity and Packaging}\label{modularity}
-Modularity is a broad term that refers to mechanisms useful for combining and reusing independently developed libraries of code.
-Languages with good support for modularity promote information hiding, thus localizing the effect of code changes, and allow modules to communicate over well-defined interfaces.
-There remains widespread disagreement and considerable ongoing research on language support for modularity.
-Object-oriented methodologies, although common in industrial projects, are used less frequently in scientific codes due to a perceived loss of control and performance \cite{basili2008understanding}.
-Many dynamic languages support modularity only implicitly using duck typing.
-Functional languages, on the other hand, often have module systems that enforce correct usage of an interface at compile-time, albeit at some notational cost \cite{tapl}.
-
-The packaging and linking mechanisms available in a language can also significantly impact its usability.
-Languages like C and C++ use a fragile preprocessor-based packaging mechanism and require separate header files, which often leads to subtle errors and compilation inefficiencies.
-More recent high-level languages have developed a varied set of mechanisms that are significantly simpler.
-
-\subsubsection{Verifiability}\label{verifiability}
-Verifying that a program does not contain errors and that it will operate according to specification (if a specification exists, which can be rare in this domain \cite{oai:open.ac.uk.OAI2:17673}) is a critical concern across all areas of software development.
-Errors in scientific programs may arise due to problems in basic program logic, as in other domains, but also due to the accumulation of  numerical approximation errors or by violation of domain-specific constraints (e.g.
-inconsistent scientific units.)
-
-A number of design decisions can influence the difficulty of formal program verification.
-Broadly stated, unconstrained support for dynamic indirection and direct access to memory have made program verification more difficult in many commonly-used languages.
-As a result, many classes of errors are only caught at run-time, often due to edge cases that are difficult to test for.
-
-Advanced type systems, matched with appropriately constrained data structures and control constructs, can dramatically increase the likelihood that an error is found at compile-time and even eliminate entire classes of errors \cite{tapl}.
-A large portion of academic research on programming language design has focused on designing such type systems.
-Unfortunately, as with many new parallel abstractions, these generally require that a language be modified with new primitives and most work has either been with prototype languages or functional languages like Ocaml, Haskell or Coq, which, due to some of the factors mentioned here, have seen  limited (though growing) adoption in scientific computing so far.
-
-In the absence of a formal proof of correctness, users must test programs with specific inputs, relying both on language-provided run-time checks and user-provided run-time assertions to catch errors.
-Most languages have unit testing frameworks designed to minimize the burden of developing unit tests, though these are used infrequently in science \cite{oai:open.ac.uk.OAI2:17673}\cite{hannay2009scientists}.
-Several languages (e.g.
-Eiffel) also have support for specifying pre- and post-conditions for functions, so that the assertions that must hold about arguments are visible in the function signature, but this feature has not been well-supported by widely-used languages to date.
-
-\subsection{Adoption Criteria}
-While most of the design criteria described in the previous sections have been the topic of significant (and ongoing) research, there are a number of other, more practical issues that can significantly affect adoption of a new language or tool.
-It should be noted that certain design decisions may make it easier to satisfy these adoption to criteria as well, an important consideration given that incentive structures in academia often do not reward efforts to improve a language along these dimensions \cite{howison2011scientific}.
-We give examples of such design decisions when we describe $\cloquence$ in Section 3.
-
-\subsubsection{Performance}
-Performance is, of course, a critical concern in scientific and (particularly \cite{basili2008understanding}) high-performance computing.
-Low-level languages typically elect to give the user direct access to hardware primitives.
-High-level languages must rely on a sophisticated compiler to translate high-level abstractions into performant code.
-Although the latter approach would be ideal, program optimization remains an active research area with many open problems.
-Indeed, many relevant algorithms have been shown to be NP-Complete, so compilers must rely on heuristics.
-Humans remain better at inventing and applying heuristics, given enough motivation and experience, than computers in many cases.
-Indeed, even when the compiler can sometimes produce better optimizations, users generally insist on being able to form a mental model of what their code is doing at the machine-level so that they can {\it reason} about the effects of code changes on overall performance \cite{squires2005programmers}.
-As such, users generally demand that low-level abstractions remain available alongside high-level abstractions.
-It may also be useful to support a model where programmers can formalize, package and directly apply optimization heuristics programmatically, rather than relying on a ``black-box'' compiler.
-
-% Squires et al, 2005 -- programmers want to know what is happening under the hood so they can reason about optimization
-
-\subsubsection{Portability}\label{portability}
-A number of surveys have revealed that portability is one of the most important issues considered by scientists and engineers \cite{hannay2009scientists}\cite{basili2008understanding}.
-Several processor architectures and operating systems are in widespread use, with more appearing on a fairly regular basis.
-Low-level languages like C have generally achieved a reasonable level of portability, although aspects of the language that are underspecified can be sources of errors.
-The OpenCL language was designed with portability as a major design criteria, and compilers for OpenCL are now available for a diverse collection of processor and accelerator architectures.
-High-level languages are generally highly portable.
-
-It should be noted, however, that simple portability is not entirely sufficient -- users want to be able to write programs that can be ported to new architectures and operating systems and achieve high performance without significant retuning.
-This remains an area of active research.
-Recent efforts to build abstractions that generate efficient code for devices with multi-level memories, as implemented in the Sequoia language \cite{fatahalian2006sequoia}, have  progressed toward this goal.
-
-\subsubsection{Useful Error Messages}\label{errors}
-When the compiler or run-time system of a language generates an error, users must determine the source of the problem using the information provided in an error message.
-Studies have shown that good error messages can dramatically reduce the time it takes to debug programs \cite{marceau2011measuring}.
-Indeed, a widely-reported frustration with C++ is that it produces overly verbose and cryptic error messages, particularly when using templates.
-
-\subsubsection{Tool and Infrastructure Support}\label{tools}
-% Squires 2005 might have a ref
-Modern software development now relies on a number of tools to assist with common tasks, such as debuggers,  profilers, syntax-aware editors, documentation generators, style checkers and interactive interpreters.
-Similarly, most established languages benefit from a centralized packaged repository (e.g.
-PyPI) and a package manager that can automate the installation of new packages.
-Such support is not trivial to develop from scratch, but few early adopters are likely willing to do without these tools for non-trivial projects \cite{squires2005programmers}.
-
-\subsubsection{Backwards Compatibility}
-End-user communities have produced a number of highly tuned and tested packages that are widely used in their fields.
-Porting these libraries to a new language requires significant effort and few communities will be willing to do so, particularly before a language is very well established.
-There is generally a greater willingness to develop wrappers that invoke functions from these packages, however.
-A powerful foreign function interface, ideally able to handle native code as well as code in existing widely-used high-level languages, enables these efforts to proceed smoothly and can provide a language with a large package library relatively early in its development.
-
-Many of the reasonably successful approaches to date are built incrementally on existing languages as libraries or as simple  extensions to existing languages (e.g.
-OpenMP, UPC, Co-array Fortran, Cilk++, MPI, CUDA, OpenCL).
-Notably, this allows existing programs to continue to function correctly, while enabling the gradual integration of the new constructs in parts of the code that would derive the greatest benefit.
-In contrast, approaches that have required total rewrites have had more difficulty recruiting early adopters (e.g.
-X10 \cite{charles2005x10}.)
-
-\subsubsection{Learning Material}
-Existing languages benefit from a large collection of books, presentations and tutorials that allow new users to get started quickly.
-The most mature languages benefit further from the availability of courses and professional training seminars.
-New languages often suffer due to the lack of such polished learning material, or in some cases, due to the lack of {\it any} significant learning material targeted at end users (rather than other language designers.)
-
-\subsubsection{Open Availability}
-New languages and abstractions are often viewed with suspicion if they do not come with source code that can be modified under a free software license.
-The most unencumbered licenses (other than public domain releases) are BSD-style licenses.
-So called {\it copyleft} licenses like the GPL are also popular, requiring primarily that modifications to the compiler be distributed under the same license.
-
-\subsubsection{Social Proof}\label{social}
-Finally, users look to the language's user community for evidence that the language is a serious effort that will not be abandoned and that there are other developers building libraries and sharing information over established communication channels.
-The availability of non-trivial demonstrations have been cited in surveys as critical to adoption as well \cite{basili2008understanding}.
-These criteria, as with many of the others described above, can be difficult to satisfy, particularly for ``clean-slate'' language designs, which may help to explain why language adoption is often slow.
-However, even projects with a narrower scope, such as new parallel or domain-specific abstractions, must tackle these issues.
-
-% What Do Programmers of Parallel Machines Need?
-% Squiers et al, 2005
-% - Global view
-% - Tool support
-% - Understanding of what the code is doing
-% 
-%As should be clear, there are many challenges facing any new language design effort.
-Many of these challenges apply also to efforts with a narrower scope (e.g.
-a new parallel or domain-specific abstraction).
-Although the research community continues to tackle open issues described as design criteria, the challenges described as adoption criteria above have not generally received as much attention.
 
@@
 \bibliographystyle{IEEEtran}
-\bibliography{research}
+\bibliography{../research}
 %

### 2012-09-04 ipdps13/criteria.tex
Abstract (may be too long?)

--- 2012-03-18 csgf11/omar-vetter-csgf11.tex
@@
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-% amspaper.tex --  LaTeX2e-based template for submissions to American 
-% Meteorological Society Journals, including
-%
-% JAS 	-- Journal of the Atmospheric Sciences
-% JAMC 	-- Journal of Applied Meteorology and Climatology
-% JPO 	-- Journal of Physical Oceanography
-% MWR 	-- Monthly Weather Review
-% JTECH -- Journal of Atmospheric and Oceanic Technology
-% WAF 	-- Weather and Forecasting
-% JCLI 	-- Journal of Climate
-% JHM 	-- Journal of Hydrometeorology
-% JAM 	-- Journal of Applied Meteorology
-%
-% Template developed by B. Papa and S. Cooley, AMS. 
-% Email questions to latex@ametsoc.org.
-%
-% August 12, 2008 (SRC)
-%	- Clarified/added header notes, comments throughout
-%	- Improved title page
-%	- Edited text of document for clarity
-%	- Altered list styles to adhere to AMS style, added comments
-%	- Removed incorrect commands (i.e., \catcode) (corrects umlaut bug)
-%	- Moved non-template commands to ametsoc.sty
-%
-% August, 2008 - B. Papa
-% - Updated to handle two column journal page output
-% - Updated text with new/modified instructions
-%
-% February, 2011 - B. Papa
-% - Updated instructions for use with EM
-%
-% March, 2011 - B. Papa
-% - Added use of \linenumbers
-%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-% PREAMBLE
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%
-% The following three commands will generate a PDF that follows all the requirements for submission
-% and peer review.  Uncomment these commands to generate this output (and comment out the two lines below.)
-%
-% DOUBLE SPACE VERSION FOR SUBMISSION TO THE AMS
-\documentclass[12pt]{article}
-\usepackage{ametsoc}
-%\linenumbers
-
-%
-% The following two commands will generate a single space, double column paper that closely
-% matches an AMS journal page.  Uncomment these commands to generate this output (and comment
-% out the two lines above. FOR AUTHOR USE ONLY. PAPERS SUBMITTED IN THIS FORMAT WILL BE RETURNED
-% TO THE AUTHOR for submission with the correct formatting.
-%
-% TWO COLUMN JOURNAL PAGE LAYOUT FOR AUTHOR USE ONLY
-%\documentclass[9pt]{article}
-%\usepackage{ametsoc2col}
-
-\input{macros}
-
-%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-% ABSTRACT
-%
-% Enter your Abstract here
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\newcommand{\myabstract}{From the perspective of computational scientists, programming systems are the most visible component of HPC systems and they serve a critical role in enabling HPC architectures that are high-performance, energy-efficient, scalable, robust, and productive.
-However, systems designed to achieve these goals must  handle increasing architectural complexity associated with hierarchical parallelism, the inertia associated with investments in legacy software and platforms and a variety of other design requirements.
-Many of these challenges are already emerging in today's multicore and heterogeneous computing systems and recent reports conclude that the road to Exascale Computing will require addressing these challenges even more directly.
-In this survey, we expand upon the design criteria that constrain programming systems for high-performance computing and highlight some emerging approaches, including recent research on general-purpose computing on graphics processors.}
-%
-\begin{document}
-%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-% TITLE
-%
-% Enter your TITLE here
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\title{\textbf{\large{Programming Systems on the Road to Exascale Computing
-}}}
-%
-% Author names, with corresponding author information. 
-% [Update and move the \thanks{...} block as appropriate.]
-%
-\author{\textsc{Cyrus Omar}\\
-\textit{\footnotesize{Carnegie Mellon University}}
-\and 
-\centerline{\textsc{Jeffrey Vetter}}\\% Add additional authors, different insitution
-\centerline{\textit{\footnotesize{Oak Ridge National Laboratory}}}
-}
-%
-% The following block of code will handle the formatting of the title page depnding on whether
-% we are formatting a double column (dc) author draft or a single column paper for submission.
-% AUTHORS SHOULD SKIP OVER THIS... There is nothing to do in this section of code.
-\ifthenelse{\boolean{dc}}
-{
-\twocolumn[
-\begin{@twocolumnfalse}
-\amstitle
-
-% Start Abstract (Enter your Abstract above.  Do not enter any text here)
-\begin{center}
-\begin{minipage}{13.0cm}
-\begin{abstract}
-	\myabstract
-	\newline
-	\begin{center}
-		\rule{38mm}{0.2mm}
-	\end{center}
-\end{abstract}
-\end{minipage}
-\end{center}
-\end{@twocolumnfalse}
-]
-}
-{
-\amstitle
-\begin{abstract}
-\myabstract
-\end{abstract}
-\newpage
-}
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-% MAIN BODY OF PAPER
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%
-\section{Introduction}
-Computer-aided simulations and data analysis techniques have transformed science and engineering.
-Surveys show that scientists and engineers now spend more than 40\% of their time writing software \cite{howison2011scientific, hannay2009scientists}.
-Most of this software targets conventional desktop hardware, but about 20\% of scientists also target either local clusters or super\-computers for more numerically-intensive computations \cite{hannay2009scientists}, a number that continues to grow as scientists face ever-larger datasets and increasingly complex systems, such as those studied in the biological, environmental and social sciences.
-
-Historically, scientists needing better performance simply purchased new hardware.
-This behavior was justified by appealing to Moore's Law, which has accurately predicted a doubling in the transistor count of integrated circuits every 18 months since the invention of the transistor in 1958 \cite{moore1998cramming}.
-Although Moore's Law remains valid, it is projected that by the end of this decade, the doubling period will increase substantially, due to fundamental physical limitations.
-More problematic, however, is that an assumed corollary of Moore's Law -- that {\it clock speeds} will also double every 18 months -- is no longer valid.
-Although transistor counts have continued to increase, clock speeds have remained flat for several years.
-It is now widely acknowledged that further performance gains will require that scientists make effective use of parallel computation.
-
-Modern high-performance computing systems provide substantial hardware support for parallel computation.
-A single node in a cluster or a single workstation may contain several general-purpose multi-core processors as well as one or more  co-processors, such as a GPU, containing hundreds of specialized compute units.
-Modern clusters in turn  contain many such nodes -- the largest supercomputers today feature hundreds of thousands of nodes.
-These machines boast peak performance on the order of $10^{15}$ floating point operations per second (1 petaFLOPS).
-By the end of the decade, researchers hope to achieve a further thousand-fold increase in peak performance, a goal termed the {\it exascale initiative} \cite{exascale-roadmap}.
-At this scale, power consumption, hardware reliability, data movement and storage infrastructure become critical issues.
-Comprehensively addressing these issues  has emerged as one of the grand challenges in modern computing, and is a major focus of ongoing research spanning a variety of disciplines, including computer science, computer architecture, applied mathematics and a number of application domains in science and engineering.
-
-In an ideal world, programmers would simply continue to write programs as they have in the past, relying on a compiler to optimize them for execution on a variety of hardware configurations at high speed.
-Unfortunately, such heroic compilers are likely impractical.
-Many problems central to automatic parallelization have been shown to be NP-Hard, or require additional information that can't be extracted from a program directly, so it is unlikely that fast, exact, fully-automatic techniques will emerge.
-Instead, optimization procedures must increasingly rely on heuristics, profiling data and domain-specific knowledge.
-Such so-called `fuzzy' approaches tend to require significant human involvement, meaning that programmers who hope to make full use of modern high-performance computing hardware must deal with complex issues beyond those they have faced in programming sequential hardware in the past.
-Recognition of this issue has led to calls to develop and improve the programming systems that developers rely on for assistance with these difficult issues.
-We use the term {\it programming systems} to encompass a broad range of tools
-, including operating system APIs, programming languages, abstractions, libraries, compilers, editing environments, verification tools, documentation tools and so on.
-
-The ultimate goal of this line of research is to maximize the end-to-end {\it productivity} of practitioners, measured by the time taken by all aspects of the process leading to a desired scientific result.
-In taking this view -- with its focus on human aspects of software development and design in addition to algorithmic and architectural considerations -- it is important to first identify and characterize the developer communities who will use these tools.
-Researchers typically classify software developers as either {\it end-user developers} or {\it professional developers}.
-End-user developers are those who have little formal training related to the tools that they use.
-These tools, as a consequence, tend to be relatively simple (spreadsheets, for example).
-Professional developers, on the other hand, have explicit training or extensive experience with software engineering practices and software development tools and techniques.
-Unfortunately, scientists and engineers do not cleanly fall into either of these groups.
-Although formal training in software development is rare \cite{oai:open.ac.uk.OAI2:17673}, scientists and engineers are more technically literate and demanding than typical end-users.
-For this reason, researchers now define a third group, {\it professional end-user developers} \cite{segal2007some}, to describe ``people working in highly technical, knowledge rich professions, such as financial mathematicians, scientists and engineers, who develop their own software in order to advance their own professional goals.''
-
-In this survey, we examine the needs of professional end-user developers who currently work with, or will in the future need to work with, high-performance computing systems, including future exascale machines.
-As we do so, we describe both existing and emerging approaches, noting where additional research and development may be needed.
-We hope that the reader will emerge with a clear view of the state of the art and develop a clear, high-level understanding of the significant challenges remaining as we move toward future high-performance, high-productivity  computing systems and applications.
-
-%\section{Programming Systems in HPC}
-%Programmers today rely on a sophisticated collection of tools and abstractions to tackle the complex challenges faced in modern software engineering projects.
-In this section, we organize these programming systems and give a brief review of the choices most of these
-
-\section{System Software}
-Systems software is the portion of the programming system typically associated with the operating system or low-level runtime environment.
-It is responsible for interacting with hardware resources, usually in a relatively transparent way to the user.
-This category includes operating system APIs, file systems and I/O libraries, and systems management software \cite{exascale-roadmap}.
-
-\subsection{Operating Systems} 
-Today, some flavor of Linux is the dominant choice in HPC machines.
-Variants of UNIX, such as AIX, and lightweight kernel operating systems such as Compute Node Kernel (CNK) are also used by some vendors (notably, IBM and Cray) \cite{hpc-survey}.
-Operating system APIs are the most primitive abstractions available to programmers, so small choices (e.g.
-in thread scheduling) can have a large impact on performance.
-The choices made to accomodate conventional desktop and server workloads may not be ideal in an HPC context, so there has been considerable interest in developing HPC-specific operating systems.
-Exascale machines introduce additional complications, because resource management will require consideration of power budgets, and hardware reliability issues will become increasingly prominent.
-In some cases, collective decisions may need to be made based on information integrated across a cluster, further complicating matters.
-
-Researchers differ about whether the simpler, more specialized lightweight kernel (LWK) operating systems dedicated to high-performance computing applications are better suited to exascale applications, outweighing the flexibility and broad developer base of systems based on a general-purpose operating system like Linux.
-In either case, however, there is a need for improved APIs for memory and thread management, performance monitoring and debugging, energy management and access to specialized hardware.
-To evaluate these APIs at scale before the associated hardware is available, whole-system simulation techniques must also be developed.
-These are all active research areas.
-
-\subsection{File Systems and I/O Libraries}
-Large, distributed file systems require specialized support at the systems level.
-A number of different high-performance file systems are in wide use today, including Lustre, GPFS and NFS.
-Together with I/O libraries such as netCDF, HDF5 and MPI-IO, these systems form the basis for data storage and retrieval on high-performance machines today \cite{hpc-survey}.
-
-Unfortunately, it is not clear that these systems can continue to scale without serious changes.
-A major component of the overall energy budget of an exascale machine will be consumed by data movement.
-Today's systems are based on traditional file systems and continue to offer strong guarantees about conflicts, snchronization and coherence.
-I/O is also generally thought of as an activity that occurs periodically (e.g.
-during checkpointing) or at the end of a simulation, rather than as an activity that occurs on an ongoing basis throughout a computation.
-These constraining requirements and assumptions make scaling a challenge, even as new storage architectures and devices emerge, such as solid state disks.
-Fortunately, many applications do not require such strong guarantees in some cases.
-As more applications interleave data analysis with computation, relieving a major burden on I/O systems, they can be targeted to more specialized tasks such as checkpointing, improving their performance.
-As such, scalable, purpose-driven, hardware-aware I/O systems that are more tightly integrated into the programming language and environment are a major focus of ongoing research.
-
-\subsection{Systems Management}
-High-performance computing systems require additional management tools to ensure that needed resources are available and divided amongst users fairly and securely.
-For example, {\it batch systems} allow users to request jobs on shared machines, and schedule these jobs in an equitable way on available resources.
-Popular choices for batch systems  include Torque, MOAB, LoadLeveler and SLURM.
-A number of other tools are also used by systems administrators to maintain the cluster's software systems, monitor hardware and ensure long-term data integrity.
-Although there are fewer scaling challenges with this aspect of the software stack than with some of the others, these tools interact with several aspects of both the software stack and the hardware, and must be updated as these aspects of the high-performance computing systems change on the road to exascale.
-
-\section{Programming Environments}
-The programming environment consists of the software that developers interact with directly to develop and debug programs.
-This category centers around programming languages and includes parallel programming abstractions, domain-specific abstractions, runtime systems, compilers, debuggers, performance analysis tools, data analysis and visualization tools, and editors \cite{exascale-roadmap}.
-
-Scientists and engineers generally prefer high-level scripting languages like MATLAB, Python, R and Perl \cite{nguyen2010survey}, particularly those bundled with powerful domain-specific libraries.
-These languages are typically interpreted rather than compiled and generally lack static type systems, relying instead on run-time (or ``dynamic'') type lookup and error handling.
-Although this may increase the flexibility of the language, these features also negatively impact performance and can lead to unexpected run-time errors that are generally impossible to rule out statically.
-This is particularly severe given that rigorous software testing practices are also quite rare in these domains \cite{nguyen2010survey}.
-When running at scale, a run-time error can be very costly, so better verification and testing of scientific code will be necessary.
-
-Along code paths where performance appears to be a bottleneck, developers today turn to statically-typed, though still unsafe, low-level languages like C and Fortran \cite{4222616}.
-These languages give users explicit control over data layout and heap allocation and require explicit type annotations on all variables.
-Although this makes writing programs more tedious which can increase the number of logic errors a programmer makes, it can also potentially improve performance, particularly if the programmer understands the target architecture well (e.g.
-cache behavior).
-While a sequential reimplementation of a critical code path using a low-level language will generally produce a modest speedup, more significant speedups on modern hardware require parallelizing over many processor cores and, on massively-parallel machines, many interconnected nodes.
-Despite well-known difficulties, low-level, unsafe  approaches that use a form of shared-memory  multithreading (e.g.
-pthreads, OpenMP) paired with explicit message passing between nodes (typically using MPI) remain widely used for these tasks as well \cite{4222616, basili2008understanding}.
-
-Although researchers often propose higher-level language features and new parallel programming abstractions that aim to strike a better balance between raw performance, productivity, code portability and verifiability (see below), end-users remain skeptical of new approaches.
-This viewpoint was perhaps most succinctly expressed by a participant interviewed in a recent study by Basili et al.
-\cite{basili2008understanding}, who stated ``I hate MPI, I hate C++.
-[But] if I had to choose again, I would probably choose the same.'' Although this sentiment is easy to dismiss as paradoxical, we believe that it demands direct examination by those in the research community working to advance the practice of scientific and high-performance computing.
-%
-% Two types of languages see widespread use: productivity-oriented and performance-oriented
-% Significant diversity in productivity-oriented languages. Valued for clean syntax and high-level abstractions. Usually very slow. Have a FFI to call into a performance-oriented language.
-% Performance-oriented languages -- C or Fortran and their derivatives.
-Direct memory access, low-level abstractions, more verbose.
-Portable (though often need to include platform-specific optimizations, which can lead to messy code).
-Not easily extended with new abstractions and primitives (witness: lots of derivative languages that are not composable because no way to implement these abstractions cleanly)
-% Generally don't support verifiability very well (performance-oriented languages have simple type systems); only exception is functional languages.
-%
-
-% Simplicity (Errors, Syntax, Core Semantics)
-% Extensibility (New Abstractions, Domain-Specific Things, Domain- and Platform-Specific Optimizations)
-% Performance (Need to allow control over data layout and movement, perhaps optionally)
-% Portability (Architectures and operating systems; support new optimizations)
-% Backwards-Compatibility (Existing codebases)
-% Verifiability (type systems, metaprogramming-based, formal semantics, small TCB)
-% Tool Support (syntax highlighting, editors, debuggers, etc.)
-% Social Proof (Will not disappear - open source, significant projects have been developed, community)
-
-As in many areas of design, it can be difficult to objectively evaluate the merit of language and tool designs.
+\section{Design and Adoption Criteria}
+The primary purpose of this paper is to describe the design of a programming language, $\cloquence$.
+Unfortunately, as in many areas of design, it can be difficult to objectively evaluate the merit of such efforts.
 To better support such evaluations, researchers in design disciplines typically develop a set of high-level  design criteria that serve as a rubric for evaluating and guiding their design efforts.
@@
 Similarly, there have been few treatments of adoption criteria for new languages and abstractions in practice, an important issue given the slow rates of adoption today.
-As such, we organize this section around a set of design and adoption criteria for new languages and abstractions based on prior empirical studies of this class of {\it professional end-user developer}, as well as our observations of characteristics common to successful projects in the past.
+We have purposefully separated these criteria from the specific language we describe in Section 3 in the hopes of starting a productive discussion within the community.
+
+\subsection{Professional End-User Developers}
+An important first step is to identify the developer communities targeted by our design efforts.
+Researchers in software engineering typically distinguish between {\it end-user developers} and {\it professional developers}.
+End-user developers have little formal training relevant to the tools that they use, and the tools themselves are relatively simple (spreadsheets, for example).
+Professional developers, on the other hand, have explicit training or extensive experience with software engineering and software development techniques.
+Scientists and engineers do not cleanly fall into either of these groups, however -- although formal training in software development is rare \cite{oai:open.ac.uk.OAI2:17673}, members of these groups are more technically literate and demanding than typical end-users.
+For this reason, researchers have proposed a third group, {\it professional end-user developers}, to describe ``people working in highly technical, knowledge rich professions, such as financial mathematicians, scientists and engineers, who develop their own software in order to advance their own professional goals.
+\cite{segal2007some}''
+As such, we consider the literature on novice and end-user programming (cf.
+\cite{pane1996usability}) in addition to that on programming language design for professional developers to justify the criteria we develop below.
+Language usability and adoption are not commonly studied topics, so many of the hypotheses we present below also rely on informal observations about languages and tools that have been successful in the past.
 
 \subsection{Design Criteria}
-We define design criteria as aspects of an abstraction or language's specification, rather than its implementation, that help developers express their intent naturally and correctly.
+We define design criteria as aspects of a language's design, rather than its implementation, that help developers express their intent naturally and correctly.
 
@@
 Cordy identifies the principle of conciseness with elimination of redundancy and the availability of reasonable defaults \cite{cordy1992hints}.
-The high-level languages listed above have all either made optional, or removed entirely, much of the syntactic overhead characteristic of low-level languages, such as explicit variable declarations and extensive headers or preambles that contain information that can be inferred from the body of the program in most cases.
-They also typically feature simple array indexing syntax and concise literal forms for common data structures like arrays, matrices, sets and maps.
+The high-level languages listed in the introduction have all either made optional, or removed entirely, much of the syntactic overhead characteristic of low-level languages, such as explicit variable declarations and extensive headers or preambles that contain information that can be inferred from the body of the program in most cases.
+They also typically feature simple array indexing syntax and concise literal forms for common data structures like arrays, sets and maps.
 A concise and minimal syntax eliminates unnecessary keystrokes and keeps more code visible on screen at a time.
@@
 
-However, the benefits of concise syntax must be balanced with concerns about readability and familiarity.
+However, one can also go too far, so these benefits must be balanced with concerns about readability and familiarity.
 A number of principles have been proposed to operationalize the notion of code readability.
@@
 Most researchers become familiar with formal notation by studying mathematics.
-In nearly all commonly used languages in scientific computing, common mathematical notations or close approximations thereof are generally used.
+In nearly all commonly used languages in scientific computing, the common mathematical notation is used whenever possible.
 In contrast, many academic languages have settled on alternative notational styles.
@@
 Additionally, a few languages, such as Mathematica, have support for more typographically-rich mathematical notation via a structural editing interface.
-This technique appears to be helpful, although we do not know of formal studies that provide evidence for this claim.
+These techniques appear to be helpful, although we do not know of formal studies that provide evidence for this claim.
 
@@
 Although library-based abstractions are useful, they often suffer from limitations of the language, particularly if they require compile-time support.
-Primitive language support for a parallel abstraction has been shown to be more useful than a library-based implementation in at least one case \cite{cave2010comparing}.
+Primitive language support for a parallel abstraction has been shown to be more usable than a library-based implementation in at least one case \cite{cave2010comparing}.
+Languages that have been designed specifically to explore a single abstraction as a core language feature often see more limited adoption than library-based implementations because they remain difficult to use in circumstances for which an alternative abstraction is more appropriate.
+Examples of abstractions where language-based implementations remain less frequently used than library-based implementations include the actor model (e.g.
+Erlang), software transactional memory \cite{harris2003language}, and global arrays (e.g.
+UPC).
 As such, it is important that a language support several modes of operation, ideally without excessively favoring one over another.
-
-Languages that have been designed specifically to explore a single abstraction as a core language feature often see limited adoption because they remain difficult to use in circumstances for which the favored abstraction is inappropriate.
-Examples of these abstractions and languages that feature them include:
-
-\begin{itemize}
-\item shared-memory concurrency (e.g. Java, C\#)
-\item share-nothing message-passing (e.g. Erlang)
-\item flat data and task parallelism (e.g. OpenMP and  OpenCL)
-\item nested data and task parallelism (e.g. NESL, Copperhead)
-\item transactional memory (e.g. the language described by Harris \cite{harris2003language})
-\item automatic parallelization of functional primitives (e.g. Data Parallel Haskell, MapReduce and others)
-\item (partitioned) global address spaces (e.g. UPC, Co-array Fortran, Global Arrays, Fortress, X10, Chapel and others)
-\item adaptive thread migration (e.g. Charm++)
-\end{itemize}
 
@@
 Template Haskell \cite{sheard2002template}) but these too have seen relatively limited development or adoption thus far.
+In Section 3, we introduce a novel static language extension mechanism that uses type-based dispatch rather than direct manipulation of syntax trees.
 
@@
 Some languages (notably Java) have taken up the philosophy that even simple language extension mechanisms should not be in the hands of end-users for this reason.
-Although this continues to be debated, we argue that extensibility is crucial for parallel and scientific programming in particular due to the significant levels of ongoing research into new parallel abstractions and the diverse set of domain-specific use cases.
+Although this continues to be debated, we argue that extensibility is crucial for parallel and scientific programming in particular due to the significant levels of ongoing research into new parallel abstractions and the diverse set of other domain-specific use cases.
 
@@
 The most recent revision of C++ now has support for closures and closures are also being considered for a future revision of Java.
-OpenCL does not come with any support for higher-order functions, even via function pointers.
+OpenCL does not come with any support for higher-order functions, an issue we explore further in Section 3.
 
@@
 Languages like C and C++ use a fragile preprocessor-based packaging mechanism and require separate header files, which often leads to subtle errors and compilation inefficiencies.
-More recent high-level languages have developed a varied set of mechanisms that are significantly simpler for packaging and deployment.
+More recent high-level languages have developed a varied set of mechanisms that are significantly simpler.
 
@@
 In the absence of a formal proof of correctness, users must test programs with specific inputs, relying both on language-provided run-time checks and user-provided run-time assertions to catch errors.
-Most languages have unit testing frameworks designed to minimize the burden of developing unit tests, though these are used infrequently in science \cite{oai:open.ac.uk.OAI2:17673, hannay2009scientists}.
+Most languages have unit testing frameworks designed to minimize the burden of developing unit tests, though these are used infrequently in science \cite{oai:open.ac.uk.OAI2:17673}\cite{hannay2009scientists}.
 Several languages (e.g.
@@
 \subsection{Adoption Criteria}
-While most of the design criteria described in the previous section have been the topic of significant (and ongoing) research, there are a number of other, more practical issues that can significantly affect adoption of a new language or tool.
-It should be noted that certain design decisions may make it easier to satisfy these adoption to criteria as well, an important consideration given that incentive structures in academia often do not reward efforts to improve a language along some of the following dimensions \cite{howison2011scientific}.
+While most of the design criteria described in the previous sections have been the topic of significant (and ongoing) research, there are a number of other, more practical issues that can significantly affect adoption of a new language or tool.
+It should be noted that certain design decisions may make it easier to satisfy these adoption to criteria as well, an important consideration given that incentive structures in academia often do not reward efforts to improve a language along these dimensions \cite{howison2011scientific}.
+We give examples of such design decisions when we describe $\cloquence$ in Section 3.
 
@@
 Several processor architectures and operating systems are in widespread use, with more appearing on a fairly regular basis.
-Low-level languages like C have generally achieved a reasonable level of portability, although aspects of the language that are underspecified can be a source of errors.
+Low-level languages like C have generally achieved a reasonable level of portability, although aspects of the language that are underspecified can be sources of errors.
 The OpenCL language was designed with portability as a major design criteria, and compilers for OpenCL are now available for a diverse collection of processor and accelerator architectures.
-High-level languages are generally highly portable due to the use of a virtual machine architecture.
+High-level languages are generally highly portable.
 
@@
 Modern software development now relies on a number of tools to assist with common tasks, such as debuggers,  profilers, syntax-aware editors, documentation generators, style checkers and interactive interpreters.
-Similarly, most established languages benefit from a centralized package repository (e.g.
-PyPI) paired with a package manager that can automate the installation of new packages.
-These tools and infrastructure are not trivial to develop from scratch, and few incentives exist for researchers to do so, but few users are willing to do without these tools for non-trivial projects \cite{squires2005programmers}.
+Similarly, most established languages benefit from a centralized packaged repository (e.g.
+PyPI) and a package manager that can automate the installation of new packages.
+Such support is not trivial to develop from scratch, but few early adopters are likely willing to do without these tools for non-trivial projects \cite{squires2005programmers}.
 
@@
 
-\subsubsection{Social Proof}
+\subsubsection{Social Proof}\label{social}
 Finally, users look to the language's user community for evidence that the language is a serious effort that will not be abandoned and that there are other developers building libraries and sharing information over established communication channels.
@@
 However, even projects with a narrower scope, such as new parallel or domain-specific abstractions, must tackle these issues.
-Funding agencies can aid in adoption by guaranteeing development will be supported for a number of years for languages that appear to have the greatest promise.
 
@@
 Although the research community continues to tackle open issues described as design criteria, the challenges described as adoption criteria above have not generally received as much attention.
-
-% An example of a floating figure using the graphicx package.
-% Note that \label must occur AFTER (or within) \caption.
-% For figures, \caption should occur after the \includegraphics.
-% Note that IEEEtran v1.7 and later has special internal code that
-% is designed to preserve the operation of \label within \caption
-% even when the captionsoff option is in effect. However, because
-% of issues like this, it may be the safest practice to put all your
-% \label just after \caption rather than within \caption{}.
-%
-% Reminder: the "draftcls" or "draftclsnofoot", not "draft", class
-% option should be used if it is desired that the figures are to be
-% displayed while in draft mode.
-%
-%\begin{figure}[!t]
-%\centering
-%\includegraphics[width=2.5in]{myfigure}
-% where an .eps filename suffix will be assumed under latex, 
-% and a .pdf suffix will be assumed for pdflatex; or what has been declared
-% via \DeclareGraphicsExtensions.
-%\caption{Simulation Results}
-%\label{fig_sim}
-%\end{figure}
-
-% Note that IEEE typically puts floats only at the top, even when this
-% results in a large percentage of a column being occupied by floats.
-
-% An example of a double column floating figure using two subfigures.
-% (The subfig.sty package must be loaded for this to work.)
-% The subfigure \label commands are set within each subfloat command, the
-% \label for the overall figure must come after \caption.
-% \hfil must be used as a separator to get equal spacing.
-% The subfigure.sty package works much the same way, except \subfigure is
-% used instead of \subfloat.
-%
-%\begin{figure*}[!t]
-%\centerline{\subfloat[Case I]\includegraphics[width=2.5in]{subfigcase1}%
-%\label{fig_first_case}}
-%\hfil
-%\subfloat[Case II]{\includegraphics[width=2.5in]{subfigcase2}%
-%\label{fig_second_case}}}
-%\caption{Simulation results}
-%\label{fig_sim}
-%\end{figure*}
-%
-% Note that often IEEE papers with subfigures do not employ subfigure
-% captions (using the optional argument to \subfloat), but instead will
-% reference/describe all of them (a), (b), etc., within the main caption.
-
-% An example of a floating table. Note that, for IEEE style tables, the 
-% \caption command should come BEFORE the table. Table text will default to
-% \footnotesize as IEEE normally uses this smaller font for tables.
-% The \label must come after \caption as always.
-%
-%\begin{table}[!t]
-%% increase table row spacing, adjust to taste
-%\renewcommand{\arraystretch}{1.3}
-% if using array.sty, it might be a good idea to tweak the value of
-% \extrarowheight as needed to properly center the text within the cells
-%\caption{An Example of a Table}
-%\label{table_example}
-%\centering
-%% Some packages, such as MDW tools, offer better commands for making tables
-%% than the plain LaTeX2e tabular which is used here.
-%\begin{tabular}{|c||c|}
-%\hline
-%One & Two\\
-%\hline
-%Three & Four\\
-%\hline
-%\end{tabular}
-%\end{table}
-
-% Note that IEEE does not put floats in the very first column - or typically
-% anywhere on the first page for that matter. Also, in-text middle ("here")
-% positioning is not used. Most IEEE journals/conferences use top floats
-% exclusively. Note that, LaTeX2e, unlike IEEE journals/conferences, places
-% footnotes above bottom floats. This can be corrected via the \fnbelowfloat
-% command of the stfloats package.
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-% REFERENCES
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-% Create a bibliography directory and place your .bib file there.
-\ifthenelse{\boolean{dc}}
-{}
-{\clearpage}
-%\bibliographystyle{ametsoc}
-%\bibliographystyle{IEEEtran}
-\bibliographystyle{plain-annote}
-\bibliography{../research}
-
-\end{document}
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-% END OF TEMPLATE
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

### 2012-09-05 ipdps13/clq-ipdps13.tex
IPDPS13: Cleaned up intro beginning

@@
 \section{Introduction}
+
+Computer-aided simulation and data analysis techniques have transformed science and engineering.
+Surveys show that scientists and engineers now spend up to 40\% of their time writing software, generally individually or in small groups \cite{howison2011scientific}\cite{hannay2009scientists}.
+Most of this software targets desktop hardware, but about 20\% of scientists also target either local clusters or super\-computers for more numerically-intensive computations \cite{hannay2009scientists}.
+To fully harness the power of modern computing resources, however, these so-called {\em professional end-user developers} \cite{segal2007some} must write parallel programs.
+
+Professional end-users today generally use high-level scripting languages like MATLAB, Python, R or Perl for tasks that are not performance-sensitive, and statically-typed, low-level languages like C and Fortran paired with low-level abstractions like pthreads and MPI for performance-critical tasks.
+These low-level languages and abstractions are notoriously tedious and difficult to use correctly, even for expert programmers.
+CITATION?
+
+The research community has proposed dozens of novel language features and parallel programming abstractions that aim to  more equitably balance important factors like performance, portability, verifiability and ease-of-use.
+Unfortunately, professional end-users have not adopted any of these {\em en masse}, and users have become  skeptical of new approaches.
+This viewpoint was perhaps most succinctly expressed by a participant interviewed in a recent study by Basili et al.
+\cite{basili2008understanding}, who stated ``I hate MPI, I hate C++.
+[But] if I had to choose again, I would probably choose the same.'' Although this sentiment is easy to dismiss as paradoxical, we believe that it demands direct examination by those in the research community working to advance the practice of scientific and high-performance computing.
+
+We begin in Section 2 by developing a set of design and adoption criteria for new languages and abstractions based on prior empirical studies of these {\it professional end-user developers} as well as our observations of characteristics common to successful projects in the past.
+
+Scientists and engineers often use high-level scripting languages like MATLAB, Python, R and Perl \cite{nguyen2010survey}, due to their powerful domain-specific libraries and their focus on ease-of-use and flexibility.
+Scripting languages are typically interpreted rather than compiled and generally do not feature static type systems, relying instead on run-time  type lookup and indirection.
+Although this can increase the flexibility of the language, these features negatively impact  performance and code verifiability.
+
+Along code paths where performance appears to be a bottleneck, professional end users typically turn to statically-typed low-level languages like C and Fortran \cite{4222616}.
+These languages are statically and explicitly typed and require users to explicitly control data layout and heap allocation.
+Although this makes writing programs more tedious and error-prone, it can significantly improve performance, particularly if attributes of the target architecture (e.g.
+caching behavior) and the problem domain are carefully considered.
+
+Although even a sequential reimplementation of a critical code path using a low-level language will produce a modest speedup, more significant speedups on modern hardware require parallelizing over many processor cores and, on massively-parallel machines, many interconnected nodes.
+Despite well-known difficulties, low-level approaches that use a form of shared-memory concurrency (e.g.
+pthreads, OpenMP) paired with explicit message passing between nodes (typically using MPI) are the most widely-used \cite{4222616}\cite{basili2008understanding}.
+
 Researchers design and implement novel parallel programming abstractions\- on a regular basis.
@@
 
-Computer-aided simulations and data analysis techniques have transformed science and engineering.
-Surveys show that scientists and engineers now spend up to 40\% of their time writing software, generally individually or in small groups \cite{howison2011scientific}\cite{hannay2009scientists}.
-Most of this software targets conventional desktop hardware, but about 20\% of scientists also target either local clusters or super\-computers for more numerically-intensive computations \cite{hannay2009scientists}.
-In recent years, GPUs and other kinds of coprocessors have emerged as powerful platforms for high-performance computing as well.
-
-Many scientists and engineers prefer high-level languages like MATLAB, Python, R and Perl \cite{nguyen2010survey}, due to their powerful domain-specific libraries and their focus on productivity and flexibility.
-These languages are typically interpreted rather than compiled and generally do not feature static type systems, relying instead on run-time (``dynamic'') type lookup and indirection.
-Although this can increase the flexibility of the language, these features also negatively impact  performance.
-Along code paths where performance appears to be a bottleneck, developers typically turn to traditional statically-typed low-level languages like C and Fortran \cite{4222616}.
-These languages require explicit type annotations on variables and give users explicit control over data layout and heap allocation.
-Although this makes writing programs more tedious and error-prone, it can also significantly  improve performance, particularly if attributes of the target architecture (e.g.
-caching behavior) are considered.
-
-While a sequential reimplementation of a critical code path using a low-level language will produce a modest speedup, more significant speedups on modern hardware require parallelizing over many processor cores and, on massively-parallel machines, many interconnected nodes.
-Despite well-known difficulties, low-level approaches that use a form of shared-memory  multithreading (e.g.
-pthreads, OpenMP) paired with explicit message passing between nodes (typically using MPI) remain widely used \cite{4222616}\cite{basili2008understanding}.
+slash
 

### 2012-09-14 ipdps13/ace-ipdps13.tex
Renamed clq to ace in file names.

--- 2012-09-05 ipdps13/clq-ipdps13.tex
@@
 
+\usepackage{color}
+
 % *** Do not adjust lengths that control margins, column widths, etc. ***
@@
 \begin{abstract}
-In this paper, we introduce the Ace compilation environment and demonstrate its practicality as a foundational tool in both high-performance computing research and in practice.
+In this paper, we introduce the Ace compilation environment and demonstrate its practicality as a foundational tool for both high-performance computing research and in practice. 
 Ace consists of a statically-typed language that shares syntax with Python and uses the Python language itself as a multi-purpose compile-time metalanguage.
-Unlike standard languages, however, the Ace compiler consists only of a minimal core that delegates control over most aspects of compilation to user libraries using a novel extension mechanism that we call {\em active type-checking and translation (\ATT)}.
-In particular, users can specify new primitive types and operations by equipping type definitions, which are first-class objects in the metalanguage, with special methods.
-The compiler then invokes these methods when type checking and translating expressions, according to a fixed type dispatch protocol associated with each syntactic form.
-Translation targets a user-defined backend language, and the safety of user-defined translations can be checked automatically.
-Using this mechanism, we implement the primitives of the C99 and OpenCL languages as simple modules within Ace.
-We go on show examples of other interesting primitive parallel  abstractions that build on top of these basic modules, demonstrating the foundational role that an actively-typed system like Ace can play in support of research on language-based parallel programming abstractions.
+Unlike standard languages, however, the Ace compiler consists only of a minimal core that delegates control over most aspects of front-end compilation to {\em active libraries} using a novel extension mechanism that we call {\em active type-checking and translation (\ATT)}.
+Users specify new primitive types and operations by equipping type definitions, which are first-class objects in the metalanguage, with designated methods that the compiler selectively invokes when type checking and translating expressions.
+Using \ATT, we implement the primitives of the C99 and OpenCL languages as modules within Ace.
+We then give examples of several first-class parallel abstractions that can be used alongside these basic primitives, demonstrating the foundational role that an actively-typed system like Ace can play for researchers developing language-based programming abstractions.
 
-In addition to \ATT, Ace supports more conventional forms of structured and unstructured metaprogramming  and integrates directly with the numeric capabilities of Python itself.
-We use these features to design a scientific simulation framework that allows users to modularly specify and orchestrate the execution of parameterized families of simulations on clusters of GPUs.
-This framework is used to successfully conduct large-scale, high-performance neuroscience simulations, providing  evidence that Ace can be used in practice today.
-
+Ace also supports more conventional forms of structured and unstructured metaprogramming and Ace programs can be compiled and launched from Python directly.
+We used these features, along with the OpenCL language module, to design a scientific simulation framework that allows users to modularly specify and orchestrate the execution of parameterized families of simulations on clusters of GPUs.
+This framework has been used to successfully conduct large-scale, high-performance neuroscience simulations, providing  evidence that Ace is also useful in practice today.
 \end{abstract}
@@
 Surveys show that scientists and engineers now spend up to 40\% of their time writing software, generally individually or in small groups \cite{howison2011scientific}\cite{hannay2009scientists}.
-Most of this software targets desktop hardware, but about 20\% of scientists also target either local clusters or super\-computers for more numerically-intensive computations \cite{hannay2009scientists}.
-To fully harness the power of modern computing resources, however, these so-called {\em professional end-user developers} \cite{segal2007some} must write parallel programs.
+Most of this software targets desktop hardware, and about 20\% of scientists also target either local clusters or super\-computers for more numerically-intensive computations \cite{hannay2009scientists}.
+To fully harness the power of modern processors, however, these so-called {\em professional end-user developers} \cite{segal2007some} must write parallel programs.
 
-Professional end-users today generally use high-level scripting languages like MATLAB, Python, R or Perl for tasks that are not performance-sensitive, and statically-typed, low-level languages like C and Fortran paired with low-level abstractions like pthreads and MPI for performance-critical tasks.
+Professional end-users today generally use high-level scripting languages like MATLAB, Python, R or Perl for tasks that are not performance-sensitive, such as simple data analysis and plotting \cite{nguyen2010survey}.
+For tasks where these interpretted, dynamically-typed languages are too slow, they turn to statically-typed, low-level languages like C and Fortran, paired with low-level abstractions like pthreads and MPI \cite{4222616}\cite{basili2008understanding}.
 These low-level languages and abstractions are notoriously tedious and difficult to use correctly, even for expert programmers.
-CITATION?
+{\color{red} CITATION?}
 

### 2012-09-17 ipdps13/ace-ipdps13.tex
Reorganized introduction to IPDPS13.

@@
 
+\usepackage[noadjust]{cite}
+% IEEE wants to use ],\,[ for this
+% but that looks dumb...
+\renewcommand{\citepunct}{,\,}
+
 % Some very useful LaTeX packages include:
@@
 %
-%\usepackage{cite}
+\usepackage{cite}
+\renewcommand{\citemid}{,~}
 % cite.sty was written by Donald Arseneau
@@
 Computer-aided simulation and data analysis techniques have transformed science and engineering.
-Surveys show that scientists and engineers now spend up to 40\% of their time writing software, generally individually or in small groups \cite{howison2011scientific}\cite{hannay2009scientists}.
+Surveys show that scientists and engineers now spend up to 40\% of their time writing software \cite{howison2011scientific, hannay2009scientists}.
 Most of this software targets desktop hardware, and about 20\% of scientists also target either local clusters or super\-computers for more numerically-intensive computations \cite{hannay2009scientists}.
-To fully harness the power of modern processors, however, these so-called {\em professional end-user developers} \cite{segal2007some} must write parallel programs.
+To fully harness the power of this modern hardware, however, these so-called {\em professional end-user developers} \cite{segal2007some} must write parallel programs.
 
-Professional end-users today generally use high-level scripting languages like MATLAB, Python, R or Perl for tasks that are not performance-sensitive, such as simple data analysis and plotting \cite{nguyen2010survey}.
-For tasks where these interpretted, dynamically-typed languages are too slow, they turn to statically-typed, low-level languages like C and Fortran, paired with low-level abstractions like pthreads and MPI \cite{4222616}\cite{basili2008understanding}.
-These low-level languages and abstractions are notoriously tedious and difficult to use correctly, even for expert programmers.
+Professional end-users today generally use high-level scripting languages like MATLAB, Python, R or Perl for tasks that are not performance-sensitive, such as small-scale data analysis and plotting \cite{nguyen2010survey}.
+For tasks where these interpretted languages are too slow, they generally call into code written in statically-typed, low-level languages like C and Fortran that uses low-level concurrency abstractions like pthreads and MPI \cite{4222616,basili2008understanding}.
+These low-level languages and abstractions are notoriously error-prone and difficult to use, even for expert programmers.
 {\color{red} CITATION?}
 
-The research community has proposed dozens of novel language features and parallel programming abstractions that aim to  more equitably balance important factors like performance, portability, verifiability and ease-of-use.
-Unfortunately, professional end-users have not adopted any of these {\em en masse}, and users have become  skeptical of new approaches.
+The research community has proposed dozens of novel language features and parallel programming abstractions that aim to  more equitably balance important concerns relating to performance, portability, verifiability and ease-of-use.
+Unfortunately, professional end-users have not adopted these {\em en masse}, and many are generally skeptical of new approaches from the research community.
 This viewpoint was perhaps most succinctly expressed by a participant interviewed in a recent study by Basili et al.
 \cite{basili2008understanding}, who stated ``I hate MPI, I hate C++.
-[But] if I had to choose again, I would probably choose the same.'' Although this sentiment is easy to dismiss as paradoxical, we believe that it demands direct examination by those in the research community working to advance the practice of scientific and high-performance computing.
+[But] if I had to choose again, I would probably choose the same.'' Although this sentiment is easy to dismiss as paradoxical, we believe that it demands direct examination by those working to advance the practice of scientific and high-performance computing.
 
-We begin in Section 2 by developing a set of design and adoption criteria for new languages and abstractions based on prior empirical studies of these {\it professional end-user developers} as well as our observations of characteristics common to successful projects in the past.
+One reason for this is that new features and abstractions tend to come in new languages.
+This is because they need first-class language and compiler support.
+Monolithic languages make it difficult to provide this.
+Thus, we need a mechanism that allows new primitives to be defined and implemented modularly and distributed with libraries.
 
-Scientists and engineers often use high-level scripting languages like MATLAB, Python, R and Perl \cite{nguyen2010survey}, due to their powerful domain-specific libraries and their focus on ease-of-use and flexibility.
-Scripting languages are typically interpreted rather than compiled and generally do not feature static type systems, relying instead on run-time  type lookup and indirection.
-Although this can increase the flexibility of the language, these features negatively impact  performance and code verifiability.
+Ace is a language-integrated compilation environment that aims to solve this problem.
+A {\em compilation environment} is a framework that allows users to programmatically control the compilation of programs.
+In Ace, each module can interleave two kinds  of program logic: compile-time and run-time.
+The compile-time logic, written in the {\em metalanguage}, is the top-level logic in each file.
+Here, users can define and import new primitive types and operators, generate functions programmatically, perform arbitrary code transformations and emit functions for later execution.
+Ace also permits a hybrid mode of use whereby the functions can also be executed immediately after being defined.
 
-Along code paths where performance appears to be a bottleneck, professional end users typically turn to statically-typed low-level languages like C and Fortran \cite{4222616}.
-These languages are statically and explicitly typed and require users to explicitly control data layout and heap allocation.
-Although this makes writing programs more tedious and error-prone, it can significantly improve performance, particularly if attributes of the target architecture (e.g.
-caching behavior) and the problem domain are carefully considered.
+Paragraph describing how active typechecking and translation work. Generalization of active libraries.
 
-Although even a sequential reimplementation of a critical code path using a low-level language will produce a modest speedup, more significant speedups on modern hardware require parallelizing over many processor cores and, on massively-parallel machines, many interconnected nodes.
-Despite well-known difficulties, low-level approaches that use a form of shared-memory concurrency (e.g.
-pthreads, OpenMP) paired with explicit message passing between nodes (typically using MPI) are the most widely-used \cite{4222616}\cite{basili2008understanding}.
+The overall design of the Ace language as a gestalt thing that satisfies a number of design and adoption criteria imposed by professional end-users.
+We demonstrate its practicality by describing an Ace-based simulator library in Section ??.
+Ace also supports researchers by allowing them to modularly specify new abstractions, which we demonstrate with examples like OpenCL and several others in Section ??.
 
-Researchers design and implement novel parallel programming abstractions\- on a regular basis.
+In addition to the general introduction of Ace as a foundational tool for research and practice and generalizable lessons relating to language design that we note below, specific novel contributions in this paper:
+
+\begin{itemize}
+\item ATT, a generalization of active libraries, where types are metalanguage objects and programs are given semantics by type-associated methods called by the compiler.
+\item A method for eliminating type annotations by using a form of type inference within functions and type propagation between functions.
+\item A method to check the correctness of generated code, by checking for satisfaction of representational consistency constraints associated with types, defined in Section ??. 
+\item Several type-aware code generation and simulation orchestration techniques used in the clegans library, described in Section ??.
+\end{itemize}
+
+%Researchers design and implement novel parallel programming abstractions\- on a regular basis.
 Each abstraction aims to make some class of developers more productive by making appropriate trade-offs between performance, portability, verifiability and ease-of-use.
@@
 Unfortunately, such languages have not been widely adopted by practitioners.
-
-slash
-
-Although researchers often propose higher-level language features and parallel programming abstractions that aim to strike a better balance between raw performance, productivity, code portability and verifiability (see Section 2), end-users remain skeptical of new approaches.
+%
+%slash
+%
+%Although researchers often propose higher-level language features and parallel programming abstractions that aim to strike a better balance between raw performance, productivity, code portability and verifiability (see Section 2), end-users remain skeptical of new approaches.
 This viewpoint was perhaps most succinctly expressed by a participant interviewed in a recent study \cite{basili2008understanding}, who stated ``I hate MPI, I hate C++.
@@
 
-We then introduce a new programming language named $\cloquence$ (pronounced ``C eloquence'') in Section 3.
+%We then introduce a new programming language named $\cloquence$ (pronounced ``C eloquence'') in Section 3.
 $\cloquence$ initially targets code paths where low-level languages, particularly OpenCL, would be used today.
 By employing a collection of powerful front-end language and compiler design techniques, making pragmatic design choices, and building on top of established infrastructure, we argue that $\cloquence$ may uniquely satisfy many of the criteria established in Section 2, particularly those related to ease of use, familiarity, performance and tool support.
+%
+%Unlike many existing languages, $\cloquence$ is designed to be highly extensible from inception, employing a powerful compile-time language extension and code generation mechanism based around the concept of {\it active libraries} \cite{activelibraries}.
+We argue that this feature may make $\cloquence$ particularly suitable as a platform for researchers developing new parallel abstractions and other specialized, high-level language features, as well as  development tools.
+%
+%In Section 4 we briefly describe a case study where the language was used to conduct large-scale spiking neural circuit simulations on GPU hardware, demonstrating the feasibility of this approach on a non-trivial problem.
+Finally, conclude in Section 6 with future directions for both the $\cloquence$ language and the research community.
+%
 
-Unlike many existing languages, $\cloquence$ is designed to be highly extensible from inception, employing a powerful compile-time language extension and code generation mechanism based around the concept of {\it active libraries} \cite{activelibraries}.
-We argue that this feature may make $\cloquence$ particularly suitable as a platform for researchers developing new parallel abstractions and other specialized, high-level language features, as well as  development tools.
+\section{ACE: An \underline{A}ctively-Typed \underline{C}ompilation \underline{E}nvironment}
+Ace is based on the Python programming language and implemented entirely as a Python package.
+The metalanguage is Python itself, the object language uses Python's grammar directly, and the compiler is implemented in Python as well.
+Write some more here.
+Why Python?
+Popularity and to aid in adoption: can be installed just like any other Python module, uses Python's module system and distribution infrastructure, piggybacks on existing Python toolchain.
 
-In Section 4 we briefly describe a case study where the language was used to conduct large-scale spiking neural circuit simulations on GPU hardware, demonstrating the feasibility of this approach on a non-trivial problem.
-Finally, conclude in Section 6 with future directions for both the $\cloquence$ language and the research community.
-
-\section{The $\cloquence$ Language}
 We now describe a programming language designed with many of the criteria described in the previous sections in mind, $\cloquence$.
@@
 
+\section{Related Work}
+
 \section{Conclusion}

### 2012-09-20 ipdps13/ace-ipdps13.tex
Cleaned up abstract.

@@
 In this paper, we introduce the Ace compilation environment and demonstrate its practicality as a foundational tool for both high-performance computing research and in practice. 
-Ace consists of a statically-typed language that shares syntax with Python and uses the Python language itself as a multi-purpose compile-time metalanguage.
-Unlike standard languages, however, the Ace compiler consists only of a minimal core that delegates control over most aspects of front-end compilation to {\em active libraries} using a novel extension mechanism that we call {\em active type-checking and translation (\ATT)}.
-Users specify new primitive types and operations by equipping type definitions, which are first-class objects in the metalanguage, with designated methods that the compiler selectively invokes when type checking and translating expressions.
-Using \ATT, we implement the primitives of the C99 and OpenCL languages as modules within Ace.
-We then give examples of several first-class parallel abstractions that can be used alongside these basic primitives, demonstrating the foundational role that an actively-typed system like Ace can play for researchers developing language-based programming abstractions.
+Ace consists of a statically-typed language that shares syntax with Python embedded within the Python language itself, which serves as the compile-time metalanguage.
+Unlike standard languages, Ace consists only of a minimal core that delegates control over most aspects of front-end compilation to {\em active libraries} using a novel extension mechanism that we call {\em active type-checking and translation (\ATT)}.
+Users specify new primitive types and operations by equipping type definitions, which are first-class objects in the metalanguage, with  methods that the compiler selectively invokes when type checking and translating expressions.
+\ATT~ allows us to  implement the primitives of the C99 and OpenCL languages as simple modules within Ace.
+We then give examples of several first-class parallel abstractions that can be defined and used alongside these basic primitives, demonstrating the foundational role that an actively-typed system like Ace can play for researchers and domain experts developing first-class abstractions as well as practitioners wishing to combine multiple such abstractions without interfacing between many distinct languages.
 
-Ace also supports more conventional forms of structured and unstructured metaprogramming and Ace programs can be compiled and launched from Python directly.
-We used these features, along with the OpenCL language module, to design a scientific simulation framework that allows users to modularly specify and orchestrate the execution of parameterized families of simulations on clusters of GPUs.
-This framework has been used to successfully conduct large-scale, high-performance neuroscience simulations, providing  evidence that Ace is also useful in practice today.
+Active library support in Ace also allows for conventional forms of structured and unstructured metaprogramming, and programs can be compiled and launched from Python directly.
+We used these features, along with the OpenCL language module and some extensions, to design a scientific simulation framework that allows users to modularly specify, compile and orchestrate the execution of parameterized families of scientific simulations on clusters of GPUs.
+This framework has been used to successfully conduct large-scale, high-performance neuroscience simulations, providing initial evidence that Ace can be used at scale in practice today.
 \end{abstract}

### 2012-09-24 ipdps13/ace-ipdps13.tex
IPDPS13: Cleaned up abstract, first couple chapters in intro.

--- 2012-09-17 ipdps13/ace-ipdps13.tex
@@
 
-\usepackage[noadjust]{cite}
+\usepackage{cite}
 % IEEE wants to use ],\,[ for this
@@
 
+%\usepackage{alltt}
+\renewcommand{\ttdefault}{txtt}
 % Some very useful LaTeX packages include:
@@
 %
-\usepackage{cite}
-\renewcommand{\citemid}{,~}
 % cite.sty was written by Donald Arseneau
@@
 \begin{abstract}
-In this paper, we introduce the Ace compilation environment and demonstrate its practicality as a foundational tool for both high-performance computing research and in practice. 
-Ace consists of a statically-typed language that shares syntax with Python and uses the Python language itself as a multi-purpose compile-time metalanguage.
-Unlike standard languages, however, the Ace compiler consists only of a minimal core that delegates control over most aspects of front-end compilation to {\em active libraries} using a novel extension mechanism that we call {\em active type-checking and translation (\ATT)}.
-Users specify new primitive types and operations by equipping type definitions, which are first-class objects in the metalanguage, with designated methods that the compiler selectively invokes when type checking and translating expressions.
-Using \ATT, we implement the primitives of the C99 and OpenCL languages as modules within Ace.
-We then give examples of several first-class parallel abstractions that can be used alongside these basic primitives, demonstrating the foundational role that an actively-typed system like Ace can play for researchers developing language-based programming abstractions.
-
-Ace also supports more conventional forms of structured and unstructured metaprogramming and Ace programs can be compiled and launched from Python directly.
-We used these features, along with the OpenCL language module, to design a scientific simulation framework that allows users to modularly specify and orchestrate the execution of parameterized families of simulations on clusters of GPUs.
-This framework has been used to successfully conduct large-scale, high-performance neuroscience simulations, providing  evidence that Ace is also useful in practice today.
+We introduce the Ace compilation environment and demonstrate its practicality as a foundational tool for both high-performance computing research and in practice. 
+Ace consists of a statically-typed programming language with user-extensible semantics controlled by a compile-time meta\-language, Python.
+Users specify new primitive types and their operations by equipping type definitions, which are first-class objects in the metalanguage, with  methods that the compiler selectively invokes when type checking and translating expressions, a mechanism we call {\em active type-checking and translation (\ATT)}.
+This mechanism allowed us to express the primitives of the C99 and OpenCL languages as simple modules within Ace.
+We then give examples of several first-class parallel abstractions that can be defined and used alongside these basic primitives, demonstrating the foundational role that an actively-typed system like Ace can play for researchers and domain experts developing first-class language abstractions, as well as practitioners wishing to use multiple such abstractions within a single program.
+Active library support in Ace also supports more general forms of type-aware metaprogramming, and Ace functions can be compiled and launched directly from Python, taking its standard numeric data structures as arguments.
+Using active library support in Ace, we designed a scientific simulation framework that allows users to modularly specify, compile and orchestrate the execution of parameterized families of scientific simulations on clusters of GPUs.
+This framework has been used to successfully conduct large-scale, high-performance neuroscience simulations, providing initial evidence that Ace can be used, at scale, in practice today.
 \end{abstract}
@@
 Surveys show that scientists and engineers now spend up to 40\% of their time writing software \cite{howison2011scientific, hannay2009scientists}.
-Most of this software targets desktop hardware, and about 20\% of scientists also target either local clusters or super\-computers for more numerically-intensive computations \cite{hannay2009scientists}.
-To fully harness the power of this modern hardware, however, these so-called {\em professional end-user developers} \cite{segal2007some} must write parallel programs.
+Most of this software targets desktop hardware and about 20\% of scientists target either local clusters or super\-computers for more numerically-intensive computations as well \cite{hannay2009scientists}.
+To fully harness the power of this modern hardware, however, these {\em professional end-user developers} \cite{segal2007some} must write parallel programs.
 
-Professional end-users today generally use high-level scripting languages like MATLAB, Python, R or Perl for tasks that are not performance-sensitive, such as small-scale data analysis and plotting \cite{nguyen2010survey}.
-For tasks where these interpretted languages are too slow, they generally call into code written in statically-typed, low-level languages like C and Fortran that uses low-level concurrency abstractions like pthreads and MPI \cite{4222616,basili2008understanding}.
-These low-level languages and abstractions are notoriously error-prone and difficult to use, even for expert programmers.
+Professional end-users today generally use high-level scripting languages like MATLAB, Python, R or Perl for tasks that are not performance-intensive, such as small-scale data analysis and plotting \cite{nguyen2010survey}.
+For tasks where these interpretted languages are too slow, they generally call into code written in a statically-typed, low-level language like C and Fortran and use low-level parallel abstractions like pthreads and MPI \cite{4222616,basili2008understanding}.
+Unfortunately, these low-level languages and abstractions are notoriously error-prone and difficult to use, even for expert programmers.
 {\color{red} CITATION?}
@@
 The research community has proposed dozens of novel language features and parallel programming abstractions that aim to  more equitably balance important concerns relating to performance, portability, verifiability and ease-of-use.
-Unfortunately, professional end-users have not adopted these {\em en masse}, and many are generally skeptical of new approaches from the research community.
+Unfortunately, professional end-users have not commonly adopted these approaches, and many have become generally skeptical of new approaches originating in the research community.
 This viewpoint was perhaps most succinctly expressed by a participant interviewed in a recent study by Basili et al.
 \cite{basili2008understanding}, who stated ``I hate MPI, I hate C++.
-[But] if I had to choose again, I would probably choose the same.'' Although this sentiment is easy to dismiss as paradoxical, we believe that it demands direct examination by those working to advance the practice of scientific and high-performance computing.
+[But] if I had to choose again, I would probably choose the same.'' Although this sentiment is easy to dismiss as paradoxical, we believe that it demands direct examination by researchers working to advance the practice of scientific and high-performance computing.
 
-One reason for this is that new features and abstractions tend to come in new languages.
-This is because they need first-class language and compiler support.
-Monolithic languages make it difficult to provide this.
+One reason for this gap is because new features and abstractions tend to be distributed within new programming languages. Prominent examples include:
+\begin{itemize}
+\item UPC, CAF, X10
+\item Erlang
+\item Cilk
+\item Charm++
+\end{itemize}
+
+This is because they need first-class language and compiler support but monolithic languages cannot be modularly and portably extended (we discuss the difference between language extensions and compiler extensions in Section XX).
+Unfortunately, writing code in multiple languages poses a host of challenges.
 Thus, we need a mechanism that allows new primitives to be defined and implemented modularly and distributed with libraries.
@@
 \begin{verbatim}
-import clq
-from clq.backends.opencl import 
-    OpenCL, get_global_id
+from ace.OpenCL import OpenCL, get_global_id
 
-@clq.fn(OpenCL)
+@OpenCL.fn
 def map(in, out, fn):
@@
     out[gid] = fn(in[gid])
-\end{verbatim}}
-\caption{Data-parallel higher-order map in $\cloquence$. The \texttt{get\_global\_id} function is an OpenCL primitive function.}
+\end{verbatim}
+}
+\vspace{-10pt}
+\caption{A generic data-parallel map function written using the user-defined Ace OpenCL module.}
 \label{map}
+\end{figure}
+\begin{figure}
+\small{
+\begin{verbatim}
+from figure1 import map
+from ace.OpenCL import global_ptr, double, sin
 
+T = global_ptr(double)
+map_sin_dbl = map.compile(T, T, sin.cl_type)
+\end{verbatim}
+}
+\vspace{-10pt}
+\caption{The generic map function compiled to map the sin function over global arrays of double-precision floating-point numbers.
+The \texttt{get\_global\_id} and \texttt{sin} functions are OpenCL primitives.}
+\label{mapsindblace}
 \end{figure}
-
 \begin{figure}
@@
 
-__kernel void map_sin_dbl(__global double *in, 
-                          __global double *out) 
+__kernel void map_sin_dbl(
+  __global double* in, 
+  __global double* out) 
 {
-    size_t gid = get_global_id(0);
+    size_t gid;
+    gid = get_global_id(0);
     out[gid] = sin(in[gid]);
+}\end{verbatim}
 }
-\end{verbatim}}
-\caption{An OpenCL kernel function that maps the sin function over a vector of double-precision floating-point numbers in the global memory address space.}
+\vspace{-10pt}
+\caption{The OpenCL kernel function generated by running \texttt{acecc figure2.py}, or by executing the  \texttt{print map\_sin\_dbl.code} statement.}
 \label{mapsinint}
+\vspace{-10pt}
 \end{figure}

### 2012-09-25 ipdps13/ace-ipdps13.tex
IPDPS13: intro

@@
 
-One reason for this gap is because new features and abstractions tend to be distributed within new programming languages. Prominent examples include:
+One reason for this gap between research and practice is because new features and abstractions are often developed and distributed together with a new programming language.
+Some prominent examples include:
 \begin{itemize}
-\item UPC, CAF, X10
-\item Erlang
-\item Cilk
-\item Charm++
+\item UPC, Fortress, X10 and Chapel, each featuring some form of first-class partitioned global address space (PGAS) support;
+\item Erlang, Charm++, Occam and Axum, featuring some form of first-class support for message-passing; 
+\item Cilk and NESL, featuring nested data parallelism
+\item {\color{red} should I flesh this out more? less?}
 \end{itemize}
 
-This is because they need first-class language and compiler support but monolithic languages cannot be modularly and portably extended (we discuss the difference between language extensions and compiler extensions in Section XX).
-Unfortunately, writing code in multiple languages poses a host of challenges.
-Thus, we need a mechanism that allows new primitives to be defined and implemented modularly and distributed with libraries.
+Unfortunately, using a domain-specific research language comes with a host of challenges.
+{\color{red} List challenges here?} Indeed, the most widely adopted parallel programming abstractions have been those that are distributed as libraries, such as MapReduce, or as simple, conservative language extensions available in existing compilers, such as OpenMP.
 
-Ace is a language-integrated compilation environment that aims to solve this problem.
+Ace is a language-integrated compilation environment that aims to make it easier for researchers to introduce new first-class language constructs as libraries.
 A {\em compilation environment} is a framework that allows users to programmatically control the compilation of programs.
-In Ace, each module can interleave two kinds  of program logic: compile-time and run-time.
-The compile-time logic, written in the {\em metalanguage}, is the top-level logic in each file.
+In Ace, users can interleave two kinds of program logic: compile-time and run-time.
+The compile-time logic, written in the {\em metalanguage}, is the top-level code in each file.
 Here, users can define and import new primitive types and operators, generate functions programmatically, perform arbitrary code transformations and emit functions for later execution.
-Ace also permits a hybrid mode of use whereby the functions can also be executed immediately after being defined.
+Ace also permits an interactive mode of use whereby the functions can also be executed immediately after being defined, a common usage scenario for professional end-users.
 
-Paragraph describing how active typechecking and translation work. Generalization of active libraries.
+Ace is an {\it actively-typed} compilation environment.
+This means that types are first-class objects in the metalanguage, and that the logic needed to typecheck and compile the operators associated with a type is defined alongside the type itself.
+The process by which this occurs is called {\em active typechecking and translation (\ATT)}, and we discuss it in detail in Section \ref{att}.
+\ATT~is flexible enough that we are able to define the entirety of the C99 and OpenCL languages   with it, demonstrating its practicality and providing a foundation for higher-level abstractions that we discuss in Section \ref{abstractions}.
+Notably, these abstractions can be imported piecemeal as libraries -- they are not defined as new languages, nor as extensions for particular compilers.
 
-The overall design of the Ace language as a gestalt thing that satisfies a number of design and adoption criteria imposed by professional end-users.
-We demonstrate its practicality by describing an Ace-based simulator library in Section ??.
-Ace also supports researchers by allowing them to modularly specify new abstractions, which we demonstrate with examples like OpenCL and several others in Section ??.
+Although the \ATT~mechanism is designed for use by researchers and domain experts, Ace is designed to be practical for professional end-user developers as well.
+It is based on the Python programming language, which is among the most broadly adopted languages in these communities.
+Indeed, the language itself is implemented entirely as a Python library -- no new executables need to be installed.
+Extension writers can target arbitrary backend languages and use existing compilers for back-end optimization.
+In this paper, we demonstrate the use of the flexible OpenCL platform for this purpose.
+Using these capabilities, we developed a
 
-In addition to the general introduction of Ace as a foundational tool for research and practice and generalizable lessons relating to language design that we note below, specific novel contributions in this paper:
+In addition to the general introduction of Ace as a foundational tool for research and practice and generalizable lessons relating to language design that we note below, we highlight several individual contributions in this paper:
 
 \begin{itemize}
-\item ATT, a generalization of active libraries, where types are metalanguage objects and programs are given semantics by type-associated methods called by the compiler.
+\item \ATT, a generalization of active libraries, where types are metalanguage objects and programs are given semantics by type-associated methods called by the compiler.
 \item A method for eliminating type annotations by using a form of type inference within functions and type propagation between functions.
-\item A method to check the correctness of generated code, by checking for satisfaction of representational consistency constraints associated with types, defined in Section ??. 
-\item Several type-aware code generation and simulation orchestration techniques used in the clegans library, described in Section ??.
+\item A method to check the correctness of generated code, by checking for satisfaction of representational consistency constraints associated with types, defined in Section \ref{repcon}. 
+\item Several type-aware code generation and simulation orchestration techniques used in the clegans library, described in Section \ref{clegans}.
 \end{itemize}

### 2012-10-12 ace/ace.tex
Renamed ipdp13 to ace

--- 2012-09-24 ipdps13/ace-ipdps13.tex
@@
 %%*************************************************************************
-
-% *** Authors should verify (and, if needed, correct) their LaTeX system  ***
-% *** with the testflow diagnostic prior to trusting their LaTeX platform ***
-% *** with production work. IEEE's font choices can trigger bugs that do  ***
-% *** not appear when using other class files.                            ***
-% The testflow support page is at:
-% http://www.michaelshell.org/tex/testflow/
-
-% Note that the a4paper option is mainly intended so that authors in
-% countries using A4 can easily print to A4 and see how their papers will
-% look in print - the typesetting of the document will not typically be
-% affected with changes in paper size (but the bottom and side margins will).
-% Use the testflow package mentioned above to verify correct handling of
-% both paper sizes by the user's LaTeX system.
-%
-% Also note that the "draftcls" or "draftclsnofoot", not "draft", option
-% should be used if it is desired that the figures are to be displayed in
-% draft mode.
-%
 \documentclass[10pt, conference, compsocconf]{IEEEtran}
-% Add the compsocconf option for Computer Society conferences.
-%
-% If IEEEtran.cls has not been installed into the LaTeX system files,
-% manually specify the path to it like:
-% \documentclass[conference]{../sty/IEEEtran}
 
 \usepackage{cite}
-% IEEE wants to use ],\,[ for this
-% but that looks dumb...
-\renewcommand{\citepunct}{,\,}
-
-%\usepackage{alltt}
+\renewcommand{\citepunct}{,\,} % IEEEtran wants to use ],\,[ for this but that looks dumb...
+
+\usepackage{alltt}
 \renewcommand{\ttdefault}{txtt}
-% Some very useful LaTeX packages include:
-% (uncomment the ones you want to load)
-
-% *** MISC UTILITY PACKAGES ***
-%
-%\usepackage{ifpdf}
-% Heiko Oberdiek's ifpdf.sty is very useful if you need conditional
-% compilation based on whether the output is pdf or dvi.
-% usage:
-% \ifpdf
-%   % pdf code
-% \else
-%   % dvi code
-% \fi
-% The latest version of ifpdf.sty can be obtained from:
-% http://www.ctan.org/tex-archive/macros/latex/contrib/oberdiek/
-% Also, note that IEEEtran.cls V1.7 and later provides a builtin
-% \ifCLASSINFOpdf conditional that works the same way.
-% When switching from latex to pdflatex and vice-versa, the compiler may
-% have to be run twice to clear warning/error messages.
-
-% *** CITATION PACKAGES ***
-%
-% cite.sty was written by Donald Arseneau
-% V1.6 and later of IEEEtran pre-defines the format of the cite.sty package
-% \cite{} output to follow that of IEEE. Loading the cite package will
-% result in citation numbers being automatically sorted and properly
-% "compressed/ranged". e.g., [1], [9], [2], [7], [5], [6] without using
-% cite.sty will become [1], [2], [5]--[7], [9] using cite.sty. cite.sty's
-% \cite will automatically add leading space, if needed. Use cite.sty's
-% noadjust option (cite.sty V3.8 and later) if you want to turn this off.
-% cite.sty is already installed on most LaTeX systems. Be sure and use
-% version 4.0 (2003-05-27) and later if using hyperref.sty. cite.sty does
-% not currently provide for hyperlinked citations.
-% The latest version can be obtained at:
-% http://www.ctan.org/tex-archive/macros/latex/contrib/cite/
-% The documentation is contained in the cite.sty file itself.
+\usepackage{color}
+\usepackage[usenames,dvipsnames,svgnames,table]{xcolor}
+\definecolor{mauve}{rgb}{0.58,0,0.82}
+\usepackage{listings}
+\lstset{
+  language=Python,
+  showstringspaces=false,
+  formfeed=\newpage,
+  tabsize=4,
+  commentstyle=\itshape,
+  basicstyle=\ttfamily\footnotesize,
+  morekeywords={lambda, self, assert},
+  numbers=left,
+  numberstyle=\footnotesize\color{gray}\textsf,
+  xleftmargin=2em,
+  stringstyle=\color{mauve}
+}
+\lstdefinestyle{Bash}{
+    language={}, 
+    moredelim=**[is][\color{blue}\bf\ttfamily]{`}{`},
+}
+\lstdefinestyle{OpenCL}{
+	language=C++,
+	morekeywords={kernel, __kernel, global, __global, size_t, get_global_id, sin}
+}
+
+\usepackage{float}
+\floatstyle{ruled}
+\newfloat{codelisting}{tp}{lop}
+\floatname{codelisting}{Listing}
 
@@
 \fi
-% graphicx was written by David Carlisle and Sebastian Rahtz. It is
-% required if you want graphics, photos, etc. graphicx.sty is already
-% installed on most LaTeX systems. The latest version and documentation can
-% be obtained at: 
-% http://www.ctan.org/tex-archive/macros/latex/required/graphics/
-% Another good source of documentation is "Using Imported Graphics in
-% LaTeX2e" by Keith Reckdahl which can be found as epslatex.ps or
-% epslatex.pdf at: http://www.ctan.org/tex-archive/info/
-%
-% latex, and pdflatex in dvi mode, support graphics in encapsulated
-% postscript (.eps) format. pdflatex in pdf mode supports graphics
-% in .pdf, .jpeg, .png and .mps (metapost) formats. Users should ensure
-% that all non-photo figures use a vector format (.eps, .pdf, .mps) and
-% not a bitmapped formats (.jpeg, .png). IEEE frowns on bitmapped formats
-% which can result in "jaggedy"/blurry rendering of lines and letters as
-% well as large increases in file sizes.
-%
-% You can find documentation about the pdfTeX application at:
-% http://www.tug.org/applications/pdftex
-
 % *** MATH PACKAGES ***
@@
 
-\usepackage{color}
+\usepackage{placeins}
 
@@
 
-\author{\IEEEauthorblockN{Cyrus Omar, Nathan Fulton, Sharschchandra Bidargaddi and Jonathan Aldrich}
+\author{\IEEEauthorblockN{Cyrus Omar, Nathan Fulton and Jonathan Aldrich}
 \IEEEauthorblockA{School of Computer Science\\
@@
 \begin{abstract}
-We introduce the Ace compilation environment and demonstrate its practicality as a foundational tool for both high-performance computing research and in practice. 
-Ace consists of a statically-typed programming language with user-extensible semantics controlled by a compile-time meta\-language, Python.
+We introduce the Ace compilation environment and demonstrate its suitability as a foundational tool for both research and practice in high-performance computing.
+Ace consists of a statically-typed programming language with {\em user-extensible semantics} defined using a compile-time meta\-language, Python.
 Users specify new primitive types and their operations by equipping type definitions, which are first-class objects in the metalanguage, with  methods that the compiler selectively invokes when type checking and translating expressions, a mechanism we call {\em active type-checking and translation (\ATT)}.
-This mechanism allowed us to express the primitives of the C99 and OpenCL languages as simple modules within Ace.
-We then give examples of several first-class parallel abstractions that can be defined and used alongside these basic primitives, demonstrating the foundational role that an actively-typed system like Ace can play for researchers and domain experts developing first-class language abstractions, as well as practitioners wishing to use multiple such abstractions within a single program.
-Active library support in Ace also supports more general forms of type-aware metaprogramming, and Ace functions can be compiled and launched directly from Python, taking its standard numeric data structures as arguments.
-Using active library support in Ace, we designed a scientific simulation framework that allows users to modularly specify, compile and orchestrate the execution of parameterized families of scientific simulations on clusters of GPUs.
-This framework has been used to successfully conduct large-scale, high-performance neuroscience simulations, providing initial evidence that Ace can be used, at scale, in practice today.
+We demonstrate the flexibility of this mechanism by implementing primitives from several widely-known languages, including the entirety of OpenCL.
+Ace also supports more general forms of metaprogramming, and Ace functions can be compiled and launched directly from Python with standard numeric data structures as arguments.
+Using each of these features, we designed a scientific simulation framework that allows users to modularly specify, compile and orchestrate the execution of parameterized families of scientific simulations on clusters of GPUs.
+This framework has been used to successfully conduct large-scale, high-performance neuroscience simulations, providing initial evidence that Ace is useful in practice today.
 \end{abstract}
 
-%\begin{IEEEkeywords}
-%programming languages, heterogeneous computing, metaprogramming, code generation
-%\end{IEEEkeywords}
+\begin{IEEEkeywords}
+programming languages, metaprogramming, type systems, extensibility, heterogeneous computing
+\end{IEEEkeywords}
 
@@
 Surveys show that scientists and engineers now spend up to 40\% of their time writing software \cite{howison2011scientific, hannay2009scientists}.
-Most of this software targets desktop hardware and about 20\% of scientists target either local clusters or super\-computers for more numerically-intensive computations as well \cite{hannay2009scientists}.
-To fully harness the power of this modern hardware, however, these {\em professional end-user developers} \cite{segal2007some} must write parallel programs.
-
-Professional end-users today generally use high-level scripting languages like MATLAB, Python, R or Perl for tasks that are not performance-intensive, such as small-scale data analysis and plotting \cite{nguyen2010survey}.
-For tasks where these interpretted languages are too slow, they generally call into code written in a statically-typed, low-level language like C and Fortran and use low-level parallel abstractions like pthreads and MPI \cite{4222616,basili2008understanding}.
+Most of this software targets desktop hardware and about 20\% of scientists also target either local clusters or super\-computers for more numerically-intensive computations \cite{hannay2009scientists}.
+To fully harness the power of these platforms, however, these {\em professional end-user developers} \cite{segal2007some} must write parallel programs.
+
+Professional end-users today generally use high-level scripting languages like MATLAB, Python, R or Perl for tasks that are not performance-sensitive, such as small-scale data analysis and plotting \cite{nguyen2010survey}.
+For portions of their analyses where these interpretted languages are too slow, they will typically call into code written in a statically-typed, low-level language like C or Fortran and use low-level parallel abstractions like pthreads or MPI \cite{4222616,basili2008understanding}.
 Unfortunately, these low-level languages and abstractions are notoriously error-prone and difficult to use, even for expert programmers.
-{\color{red} CITATION?}
-
-The research community has proposed dozens of novel language features and parallel programming abstractions that aim to  more equitably balance important concerns relating to performance, portability, verifiability and ease-of-use.
-Unfortunately, professional end-users have not commonly adopted these approaches, and many have become generally skeptical of new approaches originating in the research community.
-This viewpoint was perhaps most succinctly expressed by a participant interviewed in a recent study by Basili et al.
+
+The research community has proposed dozens of novel language features and parallel programming abstractions that aim to  more equitably balance important concerns related to performance, portability, verifiability and ease-of-use.
+Unfortunately, professional end-users are hesitant to adopt these, with many generally skeptical regarding the practicality of new approaches that originate in the research community.
+This viewpoint was perhaps most succinctly expressed by a participant in a recent study by Basili et al.
 \cite{basili2008understanding}, who stated ``I hate MPI, I hate C++.
-[But] if I had to choose again, I would probably choose the same.'' Although this sentiment is easy to dismiss as paradoxical, we believe that it demands direct examination by researchers working to advance the practice of scientific and high-performance computing.
-
-One reason for this gap is because new features and abstractions tend to be distributed within new programming languages. Prominent examples include:
+[But] if I had to choose again, I would probably choose the same.'' Although this sentiment is easy to dismiss as paradoxical, we believe that it demands direct examination by researchers working to advance the practice of high-performance computing.
+
+A major barrier to adoption of new language features and abstractions is that they are often developed and distributed together with a new programming language.
+Some prominent examples of this language-oriented approach include:
 \begin{itemize}
-\item UPC, CAF, X10
-\item Erlang
-\item Cilk
-\item Charm++
+\item UPC \cite{upc}, CAF \cite{caf}, Fortress \cite{fortress}, X10 \cite{x10} and Chapel \cite{chapel}, which support different forms of the partitioned global address space (PGAS) model; 
+\item Actors, CSPs and other forms of message passing are best supported by languages like Erlang, Charm++, Occam and Axum.
+\item Nested data parallelism is best supported by languages like Cilk and NESL.
 \end{itemize}
 
-This is because they need first-class language and compiler support but monolithic languages cannot be modularly and portably extended (we discuss the difference between language extensions and compiler extensions in Section XX).
-Unfortunately, writing code in multiple languages poses a host of challenges.
-Thus, we need a mechanism that allows new primitives to be defined and implemented modularly and distributed with libraries.
-
-Ace is a language-integrated compilation environment that aims to solve this problem.
-A {\em compilation environment} is a framework that allows users to programmatically control the compilation of programs.
-In Ace, each module can interleave two kinds  of program logic: compile-time and run-time.
-The compile-time logic, written in the {\em metalanguage}, is the top-level logic in each file.
-Here, users can define and import new primitive types and operators, generate functions programmatically, perform arbitrary code transformations and emit functions for later execution.
-Ace also permits a hybrid mode of use whereby the functions can also be executed immediately after being defined.
-
-Paragraph describing how active typechecking and translation work. Generalization of active libraries.
-
-The overall design of the Ace language as a gestalt thing that satisfies a number of design and adoption criteria imposed by professional end-users.
-We demonstrate its practicality by describing an Ace-based simulator library in Section ??.
-Ace also supports researchers by allowing them to modularly specify new abstractions, which we demonstrate with examples like OpenCL and several others in Section ??.
-
-In addition to the general introduction of Ace as a foundational tool for research and practice and generalizable lessons relating to language design that we note below, specific novel contributions in this paper:
-
-\begin{itemize}
-\item ATT, a generalization of active libraries, where types are metalanguage objects and programs are given semantics by type-associated methods called by the compiler.
-\item A method for eliminating type annotations by using a form of type inference within functions and type propagation between functions.
-\item A method to check the correctness of generated code, by checking for satisfaction of representational consistency constraints associated with types, defined in Section ??. 
-\item Several type-aware code generation and simulation orchestration techniques used in the clegans library, described in Section ??.
-\end{itemize}
+Unfortunately, using a specialized research language comes with a host of practical challenges, such as the cost of learning the language, and the lack of well-developed libraries and tools.
+In fact, the most widely adopted parallel programming abstractions have been those that are distributed as libraries, such as MapReduce, Global Arrays and MPI.
+This approach is more practical because it allows developers to progressively introduce parallelism into existing programs, leverage a familiar and well-developed general-purpose programming language when not directly invoking parallel primitives, and utilize different abstractions within a single program without encountering friction inherent to language boundaries.
+Abstractions developed as libraries are limited, however, in that they cannot alter or augment the semantics of the host language itself.
+Thus, abstractions that require new type checking rules and compilation logic for the types and associated operations they introduce are at a relative disadvantage in practice.
+
+Ace aims to solve this problem by allowing developers to define and distribute new first-class language constructs in a safely composable manner directly within user libraries using a mechanism called {\em active typechecking and translation (\ATT)}.
+In particular, \ATT~allows library authors to extend the semantics of the  language while retaining a fixed but flexible grammar.
+Aided by a form of type inference, the Ace grammar -- a large subset of Python's -- is flexible enough that we are able to define all of the primitives of the C99 and OpenCL languages in a natural manner.
+These primitives can then be supplemented with higher-level user-defined primitives, examples of which we discuss in Section \ref{abstractions}, as desired.
+Notably, these primitives are imported via the normal library loading mechanism; they are not defined within domain-specific languages, nor as extensions for a particular compiler for a language.
+
+%To enable this and other, Ace programs contain both compile-time and run-time program logic.
+The compile-time logic, written in the {\em metalanguage}, is the top-level code in each file.
+Types are first-class objects in the metalanguage, and the logic needed to typecheck and compile the operators associated with a type is bundled with the type definition itself.
+Abstractions can target existing backend languages, using its existing compilation infrastructure for low-level optimizations, so only the high-level semantics need to be encoded by researchers.
+Consistent with its design intent, we demonstrate the use of OpenCL as a flexible backend language and compiler framework \cite{opencl}.
+%Here, users can define and import new primitive types and operators, as well as do other types of metaprogramming tasks like generate functions programmatically, perform arbitrary code transformations.
+% Ace also permits an interactive mode of use whereby the functions can also be executed immediately after being defined, a common usage scenario for professional end-users.
+
+Although the \ATT~mechanism, a key feature of Ace, is designed for use by researchers and domain experts, Ace is designed to be highly practical for professional end-user developers as well.
+It is based on the Python programming language, which enjoys broad adoption in this community.
+Indeed, the language itself is implemented entirely as a Python library, so many existing tools, such as editor modes, documentation tools and style checkers, are available immediately.
+Ace programs can be generated programmatically and called directly from Python, with numeric arrays as arguments, so Ace integrates immediately with existing workflows to speed up performance-critical portions of code that may benefit from static compilation and support for parallelism.
+We demonstrate several of these features by developing a scientific simulation framework, which we use for specifying and orchestrating families of large-scale neurobiological circuit simulations on clusters of GPUs, in Section \ref{clegans}.
 
@@
 
-%We then introduce a new programming language named $\cloquence$ (pronounced ``C eloquence'') in Section 3.
-$\cloquence$ initially targets code paths where low-level languages, particularly OpenCL, would be used today.
-By employing a collection of powerful front-end language and compiler design techniques, making pragmatic design choices, and building on top of established infrastructure, we argue that $\cloquence$ may uniquely satisfy many of the criteria established in Section 2, particularly those related to ease of use, familiarity, performance and tool support.
-%
-%Unlike many existing languages, $\cloquence$ is designed to be highly extensible from inception, employing a powerful compile-time language extension and code generation mechanism based around the concept of {\it active libraries} \cite{activelibraries}.
-We argue that this feature may make $\cloquence$ particularly suitable as a platform for researchers developing new parallel abstractions and other specialized, high-level language features, as well as  development tools.
+%We then introduce a new programming language named Ace (pronounced ``C eloquence'') in Section 3.
+Ace initially targets code paths where low-level languages, particularly OpenCL, would be used today.
+By employing a collection of powerful front-end language and compiler design techniques, making pragmatic design choices, and building on top of established infrastructure, we argue that Ace may uniquely satisfy many of the criteria established in Section 2, particularly those related to ease of use, familiarity, performance and tool support.
+%
+%Unlike many existing languages, Ace is designed to be highly extensible from inception, employing a powerful compile-time language extension and code generation mechanism based around the concept of {\it active libraries} \cite{activelibraries}.
+We argue that this feature may make Ace particularly suitable as a platform for researchers developing new parallel abstractions and other specialized, high-level language features, as well as  development tools.
 %
 %In Section 4 we briefly describe a case study where the language was used to conduct large-scale spiking neural circuit simulations on GPU hardware, demonstrating the feasibility of this approach on a non-trivial problem.
-Finally, conclude in Section 6 with future directions for both the $\cloquence$ language and the research community.
-%
-
-\section{ACE: An \underline{A}ctively-Typed \underline{C}ompilation \underline{E}nvironment}
-Ace is based on the Python programming language and implemented entirely as a Python package.
-The metalanguage is Python itself, the object language uses Python's grammar directly, and the compiler is implemented in Python as well.
-Write some more here.
-Why Python?
-Popularity and to aid in adoption: can be installed just like any other Python module, uses Python's module system and distribution infrastructure, piggybacks on existing Python toolchain.
-
-We now describe a programming language designed with many of the criteria described in the previous sections in mind, $\cloquence$.
-$\cloquence$ is built around a minimal core and nearly every construct in the language is implemented as a library.
-Rather than beginning directly with this extension mechanism, however, let us begin with a concrete example of the language in use.
-
-\subsection{Example: Higher-Order Map for OpenCL}
-Figure \ref{map} shows the standard data-parallel map function written using an extension that implements the full OpenCL language as a library.
-Each thread, indexed by \verb|gid|, applies the transfer function, \verb|fn|, to the  corresponding element of the input array, \verb|in|, writing the result into the corresponding location in the output array, \verb|out|.
-
-\subsection{Syntax}
-Readers familiar with the Python programming language will recognize the style of syntax used in Figure \ref{map}.
-In fact, $\cloquence$ uses the Python grammar and parsing facilities directly.
-Several factors motivated this design decision.
-First, Python's syntax is widely credited as being particularly simple and readable, due to its use of significant whitespace and conventional mathematical notation.
-Python is one of the most widely-used languages in scientific computing, so its syntax is already familiar to much of the field.
-And significantly, a large ecosystem of tools already exist that work with Python files, such as code editors, syntax highlighters, style checkers and documentation generators.
-These can be used without modification to work with $\cloquence$ files.
-Therefore, by re-using an existing, widely-used grammar, we are able to satisfy many of the design criteria described in Section \ref{syntax} and the adoption criteria described in Section \ref{tools} without significant development effort.
-
-\begin{figure}
-\small{
-\begin{verbatim}
+Finally, conclude in Section 6 with future directions for both the Ace language and the research community.
+%
+
+\begin{codelisting}
+\begin{lstlisting}
+from ace.OpenCL import OpenCL, printf
+
+print "Hello, compile-time world!"
+
+@OpenCL.fn
+def main():
+    printf("Hello, run-time world!")
+main = main.compile()
+
+print "Goodbye, compile-time world!"
+\end{lstlisting}
+\caption{\texttt{hello.py}}
+\label{hello}
+\end{codelisting}
+\begin{codelisting}
+\begin{lstlisting}[style=Bash]
+$ `acecc hello.py`
+Hello, compile-time world.
+Goodbye, compile-time world.
+$ `cat hello.cl`
+__kernel void main() {
+    printf("Hello, run-time world.");
+}\end{lstlisting}
+\caption{Compiling \texttt{hello.py} using \texttt{acecc}.}
+\label{helloout}
+\end{codelisting}
+
+\section{Ace for Professional End-Users}
+%Ace is an {\em actively-typed, language-integrated compilation environment}. Let us begin by defining each of these terms. 
+%A {\em compilation environment} is a system that allows developers to programmatically control aspects of compilation directly.
+The language used to do so is called the {\em metalanguage} while the grammar that programs themselves are written in is called the {\em source grammar}.
+The metalanguage of Ace is Python and the source grammar is a significant subset of Python's own grammar.
+Programs and metalanguage code can be interleaved within the same file, so Ace is {\em language-integrated}.
+Ace programs are statically-typed, and types themselves are objects in the metalanguage.
+They can be equipped by developers with methods that control how operators associated with them are typechecked and compiled.
+Because of the relationship between this mechanism and the idea of {\em active libraries} \cite{activelibraries}, discussed further in Section \ref{related}, we call them {\em active types}, and thus Ace is {\em actively-typed}.
+\subsection{Example 1: Hello, World!}
+To demonstrate the structure of programs and libraries in Ace, we begin in Listing \ref{hello} with a variant of the canonical ``Hello, World!'' example written using the OpenCL module distributed with Ace.
+We note that the OpenCL module, which we use throughout this paper in examples, is simply a library like any other.
+The core of Ace gives no special treatment to it; it is distributed with Ace for convenience.
+
+Perhaps the most immediately apparent departure from the standard ``Hello, World!'' example comes on lines 3 and 10 of Listing \ref{hello}, which contain \verb|print| statements that are executed at {\em compile-time}.
+Indeed, Ace programs are Python scripts at the top-level, rather than a list of declarations as in most conventional languages.
+In other words, Python is the {\em compile-time metalanguage} of Ace.
+In other words, Ace programs are embedded within Python.
+By making this choice, Ace can  use Python's well-developed module system and associated distribution infrastructure directly (Line 1).
+
+The \verb|main| function defined on Lines 5-7 contains a single call to the \verb|printf| primitive imported from the \verb|ace.OpenCL| module on Line 1.
+The function is introduced using the \verb|@OpenCL.fn| decorator, which indicates that it is a {\em actively-typed} Ace function containing run-time logic written for the \verb|OpenCL| backend, also imported on Line 1.
+Without this decorator, the function would simply be a conventional Python function that could be called only at compile-time (doing so would fail here: \verb|printf| is not a Python primitive.)
+
+On Line 8, the {\em generic function} defined on Lines 5-7, is compiled to produce a {\em concrete function} which we also call \verb|main| (overwriting the previous definition).
+The distinction between generic and concrete functions will be explained in our next example.
+For now, we note that only concrete functions are exported when running the Ace compiler, called \verb|acecc|, shown operating at the shell in Listing \ref{helloout}.
+The \verb|acecc| compiler operates as follows:
+\begin{enumerate}
+\item Executes the provided Python module (\verb|hello.py|)
+\item Produces source code associated with any concrete functions (functions produced using the \verb|compile| method on Line 8 of Listing \ref{hello}) that are in the top-level environment.
+Each backend may produce one or more files (here, only \verb|hello.cl| is produced).
+\end{enumerate}
+
+Use of the \verb|acecc| executable is simply a convenience; the generated code is also available as an attribute of the concrete function: \verb|main.code|.
+We will also show in Section \ref{pyopencl} that, for backend languages with Python bindings, such as OpenCL and C, functions can be executed directly, without an intermediate compilation step, if desired.
+
+\subsection{Example 2: Higher-Order Map for OpenCL}
+\begin{codelisting}
+\begin{lstlisting}
 from ace.OpenCL import OpenCL, get_global_id
@@
 @OpenCL.fn
-def map(in, out, fn):
+def map(input, output, f):
     gid = get_global_id(0)
-    out[gid] = fn(in[gid])
-\end{verbatim}
+    output[gid] = f(input[gid])
+\end{lstlisting}
+\caption{[\texttt{listing3.py}] A generic data-parallel higher-order map function written using the user-defined Ace OpenCL module.}
+\label{map}
+\end{codelisting}
+
+\begin{codelisting}
+\begin{lstlisting}
+from listing3 import map
+from ace.OpenCL import global_ptr, double, int
+
+@OpenCL.fn
+def add5(x):
+    return x + 5
+
+A, B = global_ptr(double), global_ptr(int)
+map_add5_dbl = map.compile(A, A, add5.ace_type)
+map_add5_int = map.compile(B, B, add5.ace_type)
+\end{lstlisting}
+\caption{[\texttt{listing4.py}] The generic map function compiled to map the sin function over global arrays of double-precision floating-point numbers.
+The \texttt{get\_global\_id} and \texttt{sin} functions are OpenCL primitives.}
+\label{mapadd5dbl}
+\end{codelisting}
+
+\begin{codelisting}
+\begin{lstlisting}[style=OpenCL]
+#pragma OPENCL EXTENSION cl_khr_fp64 : enable
+
+double add5(double x) {
+    return x + 5;
 }
-\vspace{-10pt}
-\caption{A generic data-parallel map function written using the user-defined Ace OpenCL module.}
-\label{map}
-\end{figure}
-\begin{figure}
-\small{
-\begin{verbatim}
-from figure1 import map
-from ace.OpenCL import global_ptr, double, sin
-
-T = global_ptr(double)
-map_sin_dbl = map.compile(T, T, sin.cl_type)
-\end{verbatim}
-}
-\vspace{-10pt}
-\caption{The generic map function compiled to map the sin function over global arrays of double-precision floating-point numbers.
-The \texttt{get\_global\_id} and \texttt{sin} functions are OpenCL primitives.}
-\label{mapsindblace}
-\end{figure}
-\begin{figure}
-\small{\begin{verbatim}
-#pragma OPENCL EXTENSION cl_khr_fp64 : enable
-
-__kernel void map_sin_dbl(
-  __global double* in, 
-  __global double* out) 
+
+__kernel void map_add5_dbl(
+  __global double* input, 
+  __global double* output) 
 {
@@
     gid = get_global_id(0);
-    out[gid] = sin(in[gid]);
-}\end{verbatim}
+    output[gid] = add5(input[gid]);
 }
-\vspace{-10pt}
-\caption{The OpenCL kernel function generated by running \texttt{acecc figure2.py}, or by executing the  \texttt{print map\_sin\_dbl.code} statement.}
-\label{mapsinint}
-\vspace{-10pt}
-\end{figure}
-
-\subsection{Semantics}
-$\cloquence$ is a compiled, statically-typed language.
-The extension used in Figure \ref{map} implements the full OpenCL type system, which includes features inherited from C99 like pointer arithmetic and OpenCL-specific constructs like vector types.
-In fact, compilation of Figure \ref{map} will produce code semantically equivalent, and with the same performance profile, as manually-written OpenCL code.
-Compilation produces OpenCL source code, so $\cloquence$, when used with this extension, inherits OpenCL's portability profile as well, as described in Section \ref{portability}.
-
-In OpenCL itself, however, there is no way to write a function like \verb|map|.
-As described in Section \ref{hof}, OpenCL lacks support for higher-order functions or polymorphism (over the types of \verb|in| and \verb|out| in this case.) To map a function over an array, developers must create a separate function for every combination of argument types and for every transfer function.
-Figure 2 shows an OpenCL kernel that implements a map specifically for the \verb|sin| function applied to arrays of doubles in global memory.
-Note that because OpenCL does not support templates or function pointers, this is the most general solution possible without explicit code generation.
-However, even in CUDA, which does support templates, there is greater syntactic overhead for specializing a function with particular types than in $\cloquence$, where it happens implicitly.
-
-To support this much more concise form for specifying kernels, $\cloquence$ uses techniques pioneered by functional programming languages. We briefly describe these below.
-
-\subsubsection{Type, Kernel and Extension Inference}
-As described in Section \ref{syntax}, professional end-users tend to prefer languages that do not require explicit type annotations for variables.
-These languages accomplish this by using dynamic typing, so that types are associated with values instead.
-Statically-typed functional programming languages have also eliminated type annotations in many cases, however, using a technique known as {\it type inference} \cite{tapl}.
-
-Type inference is a technique for giving types to variables by examining how these variables are being used.
-As a simple example, in Figure \ref{map}, the variable \verb|gid| is being used to hold the result of calling the \verb|get_global_id| function so the type of \verb|gid| must be consistent with the return type of that function, \verb|size_t| (a device-dependent integer type.) Note, however, that in C99, many possible integer types can be given to \verb|gid| (in practice, a 32-bit integer is often used.) Picking a specific type may require further examining how \verb|gid| is used in the remainder of the function.
-To deal with this non-locality, type inference is typically cast as a constraint solving problem -- the type of each program variable is treated as an unknown and each use of the variable introduces a new constraint.
-A constraint-solving algorithm, generally a variant of unification, can be used to find a type assignment that satisfying all constraints.
-
-$\cloquence$ uses a variant of this approach (modified to admit extensions to the type system, as we will discuss shortly) to conduct whole-function type inference.
-This approach differs from the HM algorithm, however, which conducts whole-program (or whole-file) inference and assigns a single type to function arguments.
-In $\cloquence$, types are {\it propagated} to functions from their calling sites (to enable structural polymorphism, described below) and inference occurs only within a function body.
-
+
+int add5__1(int x) {
+    return x + 5;
+}
+
+__kernel void map_add5_int(
+  __global int* input,
+  __global int* output)
+{
+    size_t = gid;
+    gid = get_global_id(0);
+    output[gid] = add5__1(input[gid]);
+}
+\end{lstlisting}
+\caption{The OpenCL code generated by running \texttt{acecc listing4.py}, or by executing the  \texttt{print listing4.map\_sin\_dbl.code} statement.}
+\label{mapout}
+\end{codelisting}
+
+The ``Hello, World!'' example demonstrates the basic structure of Ace programs, but it does not require working with types.
+Listing \ref{map} shows a data-parallel map primitive written using the OpenCL library, introduced above.
+OpenCL is a data-parallel language -- users can define functions, called {\em kernels}, that execute across thousands of threads.
+Each kernel has access to its index, called its {\em global id}, which can be used by the programmer to ensure that each thread operates on different parts of the input data.
+The \verb|map| kernel defined in Listing \ref{map} applies a transfer function, \verb|f|, to the element of the input array, \verb|input| corresponding to its global id, writing the result into the corresponding location in the provided output array, \verb|output|.
+
+As written, \verb|map| is a {\em generic function} (an instance of the class \verb|ace.GenericFn|).
+This means that it does not have types assigned to its arguments.
+Indeed, the functionality of \verb|map| can be defined for many combinations of array types for \verb|input| and \verb|output| and functions.
+In this sense, \verb|map| is actually a family of functions defined for all types assigned to \verb|input|, \verb|output| and \verb|f| for which the operations in the function's body are typesafe.
+Running \verb|acecc listing3.py| would produce no output, however.
+
+To create a {\em concrete function} (an instance of the class \verb|ace.ConcreteFn|) that can be emitted by the compiler,  types must be assigned to each of the arguments of any externally callable functions.
+Listing \ref{mapadd5dbl} shows how to use the \verb|compile| method to specialize \verb|map| to apply the \verb|add5| function, defined on Lines 4-6, to arrays  that reside in global memory (OpenCL has a notion of four different memory spaces).
+Line 9 produces a version specialized for arrays of \verb|double|s and Line 10 produces a version for arrays of \verb|int|s.
+The output of compilation is shown in Figure \ref{mapout}.
+
+\subsection{Types as Metalanguage Objects}
+The types of the arguments of the functions created on Lines 4.9 and 4.10 were written in pure OpenCL would be \verb|__global double*| and \verb|__global int*| respectively.
+The corresponding types in Ace are \verb|global_ptr(double)| and \verb|global_ptr(int)|, abbreviated as \verb|A| and \verb|B| respectively on Line 8 for convenience (note that here, standard variable assignment in the metalanguage is serving the role that a \verb|typedef| declaration would in a C-like language.)
+
+Types in Ace are values in the metalanguage, Python.
+More specifically, types are {\em object instances} of user-defined classes that inherit from \verb|ace.Type|.
+For example, \verb|global_ptr(double)| is an instance of \verb|ace.OpenCL.GlobalPtrType| instantiated with the target type, \verb|double|, as an argument.
+The type \verb|double| itself (imported from \verb|ace.OpenCL| on Line 4.2) is an instance of \verb|ace.OpenCL.FloatType|.
+As we will describe further in Section \ref{att}, this notion of ``types as metalanguage objects'' is a key to Ace's flexibility.
+
+\subsection{Type Propagation and Higher-Order Functions}
+The type assigned to the third argument, \verb|f|, on Lines 9 and 10, is given as \verb|add5.ace_type|.
+The \verb|ace_type| attribute of a generic function is an instance of \verb|ace.GenericFnType|, the type of Ace generic functions, that corresponds to that particular function, \verb|add5| in this case.
+Generic functions are compiled automatically when they are called.
+That is, when the compiler encounters the call to \verb|f| inside \verb|map| when compiling \verb|map_add5_double|, it compiles a version of \verb|add5| specialized to the \verb|double| type, and similarly with the \verb|int| type when compiling \verb|map_add5_int|.
+This mechanism is called {\em type propagation}.
+In other words, we did not need to run \verb|add5.compile(double)| before compiling \verb|map_add5_dbl| (although doing so would also be valid).
+Only functions that are never called in the process of compiling other functions in a concrete module need type information explicitly provided, whereas \verb|add5| is a function that is only called within \verb|map|.
+
+We note that this scheme allows for a form of higher-order functional programming even when targeting languages, like OpenCL, that have no support for higher-order functions (OpenCL, unlike C99, does not support function pointers.) This works because the \verb|ace.GenericFnType| for one function, such as \verb|add5|, is not the same as the \verb|ace.GenericFnType| for a superficially similar function, such as \verb|add6| (defined as you would expect.) To put it in type theoretic terms, \verb|ace.GenericFnType|s are {\em uniquely inhabited} by a single generic function (and similarly for \verb|ace.ConcreteFnType|s), and thus it is impractical to use them as full first-class values (i.e.
+they cannot put into a run-time array.) In practice, this limitation is rarely a problem, particularly in parallel programming where such forms of specialization are already common in order to avoid the potential performance penalty associated with a function pointer dereference per call.
+
+We note that fully-featured first-class functions can be implemented when targeting a backend that supports them, such as C99, or by using a jump table implementation, but we do not discuss this further due to lack of space.
+%%% TODO: Refer to line numbers in output
+
+\subsection{Type Inference}
+Within the \verb|map| function in Listing \ref{map}, on Line 5, the variable \verb|gid| is initialized with the result of calling the \verb|get_global_id| primitive (OpenCL supports multidimensional global ids, so the argument specifies which dimension to use.)  Note that a type for the \verb|gid| variable is not listed explicitly.
+This is because Ace supports a form of within-function {\em type inference}.
+In this case, \verb|gid| will have type \verb|size_t| because that is the return type of \verb|get_global_id| (as defined in the OpenCL specification, which the \verb|ace.OpenCL| module follows.) This can be seen on Lines 11 and 24 in Listing \ref{mapout}.
+
+Inference is not restricted to single assignments, as in the \verb|map| example.
+Multiple assignments to the same variable with values of differing types and multiple return statements can be unified, such that the variable or return type is given a common supertype.
+For example, in Listing \ref{inference} the variable \verb|y| in the first branch of the conditional is assigned the \verb|int| literal \verb|0|.
+However, in the second branch of the loop, its type depends on the types of \verb|x| and \verb|scale|.
+We show two examples where this is the case on Lines 10 and 11.
+However, type inference correctly unifies these two types according to the C99 rules governing numeric types (which can be defined by the user, as we show in Section \ref{att}).
+This example would also work correctly if the assignments to \verb|y| were replaced with \verb|return| statements were placed directly in the branches of the conditional.
+
+\begin{codelisting}
+\begin{lstlisting}
+from ace.OpenCL import OpenCL, int, double, long
+
+@OpenCL.fn
+def threshold_scale(x, scale):
+    if x <= 0:
+        y = 0
+    else:
+        y = scale * x
+    return y
+f = f.compile(int, double)
+g = g.compile(int, long)
+assert f.return_type == double
+assert g.return_type == long
+\end{lstlisting}
+\caption{A function demonstrating whole-function type inference when multiple values with differing types are assigned to a single variable.}
+\label{inference}
+\end{codelisting}
+
+\subsection{Attribute and Extension Inference}
 In addition to type annotations, OpenCL introduces additional syntactic burdens.
-Developers must annotate functions that meet the requirements to be called from the host with the \verb|__kernel| attribute, and several types (notably, \verb|double|) and specialized functions require that an OpenCL extension be enabled with a \verb|#pragma| (as  can be seen in Figure \ref{mapsinint}.) The OpenCL extension we have developed automatically infers many of these annotations as well, consistent with the principle of conciseness described in Section \ref{syntax}.
-
-\subsubsection{Structural Polymorphism}
-In Section \ref{hof}, we discussed several strategies for achieving {\it polymorphism} -- the ability to create functions and data structures that operate over more than a single type.
-In $\cloquence$, all functions are implicitly polymorphic and can be called with arguments of {\it any type that supports the operations used by the function}.
-For example, in Figure \ref{map}, \verb|in| can be any type that supports indexing by a variable of type \verb|size_t| to produce a value of a type that can be passed into \verb|fn|, which must then produce a value consistent with indexing into \verb|out|.
-OpenCL pointer types are consistent with these constraints, for example.
-Although powerful, this also demonstrates a caveat of this approach -- that it is more difficult to give a function a concise signature, because arguments are constrained by capability, rather than to a single type \cite{malayeri2009structural}.
-
-Structural typing can be compared to the approach taken by dynamically-typed languages that rely on ``duck typing''.
-It is more flexible than the parametric polymorphism found in many functional languages and in languages like Java (which only allow polymorphic functions that are valid for {\it all} possible types), but is of comparable strength to the template system found in C++.
-It can be helpful to think of each function as being preceded by an implicit template header that assigns each argument its own unique type parameter.
-At function call sites, these parameters are implicitly specialized with the types of the provided arguments.
-This choice is again motivated by the criteria of conciseness given in Section \ref{syntax}.
-
-\subsubsection{Higher-Order Functions}
-Section \ref{hof} also discusses the value of higher-order functions.
-The OpenCL extension supports higher-order functions, despite the absence of function pointers from OpenCL itself, by explicitly passing functions as constants at compile-time.
-In other words, each function {\it uniquely inhabits} a type corresponding to it.
-
-Functions are not, however, {\it first-class} -- references to functions cannot be stored in memory, for example.
-Most use cases that we have considered thus far have never needed to do so, however.
-Compared to the function pointer approach used in other C-based languages, this approach is more efficient.
-It can, however, result in ``code bloat'', since a new copy of the higher-order function is created for each unique function argument passed to it.
-First-class functions can theoretically be simulated using an auxiliary lookup function in OpenCL, and we leave clean support for this in $\cloquence$ as open future work.
-
-\subsection{Active Libraries in $\cloquence$}
-In addition to adopting Python's grammar, $\cloquence$ uses the Python language itself as a {\it compile-time metalanguage}.
-Every $\cloquence$ file is, at the top-level, a Python script.
-This script is executed during compilation and has access to a number of powerful capabilities that allow developers to influence the compilation process and manipulate programs directly.
-Libraries that have such capabilities has been called {\it active libraries} in prior proposals \cite{activelibraries}.
-A number of  projects, such as Blitz++, have taken advantage of the C++ preprocessor and template-based metaprogramming system to implement domain-specific optimizations.
-In $\cloquence$, we replace these brittle mini-languages with a general-purpose language.
-This allows for several interesting uses that we discuss in the following sections.
-
-\subsection{Module System}
-In Figure \ref{map}, the only compile-time operations are library imports and a declaration of a $\cloquence$ function.
-However, this already demonstrates an important benefit of this approach: $\cloquence$ uses Python's simple but flexible module system.
-As a consequence, libraries written using $\cloquence$ can directly utilize the module distribution infrastructure available for Python, partially relieving the issues discussed in Section \ref{modularity}.
-
-\subsection{Compilation and Invocation}
-$\cloquence$ functions can be invoked in one of two ways: from any host language supporting OpenCL by using the standalone compiler, or directly from Python itself. 
-\begin{figure}\small{\begin{verbatim}
-from figure1 import map
-from clq.backends.opencl import double, sin
-
-map_sin_double = map.specialize(
-    double.global_ptr, double.global_ptr, 
-    sin.clq_type)
-\end{verbatim}}
-\caption{Programmatically specializing the \texttt{map} function from Figure \ref{map} for use by the standalone compiler to produce code equivalent to Figure \ref{mapsinint}.}
-\label{specialization}
-\end{figure}
-
-\subsubsection{Programmatic Specialization and Compilation}
-The standalone $\cloquence$ compiler, \verb|clqcc|, generates source code from $\cloquence$ source files.
-To do so, however, users must {\it specialize} any {\it externally callable} functions with their specific types.
-Figure \ref{specialization} shows how the specific OpenCL function given in Figure \ref{mapsinint} can be recovered from the generic specification of \verb|map| given in Figure \ref{map}.
-The \verb|specialize| method of a generic function like \verb|map| creates a new function that is concrete -- that is, with concrete argument types.
-
-The standalone compiler operates by first executing the module, then iterating over the externally available variables in the module's environment to extract the OpenCL functions and any other functions and constructs that they depend on.
-This allows developers who wish to use the host language of their choice, rather than Python, to continue writing library functions generically, in one place, only writing small stub modules to specialize them as needed.
-
-\subsubsection{Direct Invocation from Python}\label{direct}
-As discussed in the introduction, a common usage scenario in scientific computing involves the use of a dynamic, high-level language for workflow orchestration and exploratory visualization and analysis, paired with a low-level language for performance-critical sections.
-Although the extension mechanism we describe shortly will permit the development of a broad array high-level language features in the future, current extensions, such as the OpenCL extension we have been describing, have been designed to make low-level programming simpler.
-For this reason, the OpenCL extension includes support for directly invoking $\cloquence$ functions from Python scripts with minimal overhead.
-
-This feature builds on top of an established library, \verb|pyopencl|, that wraps the OpenCL host API and integrates it with the popular \verb|numpy| numerics library in Python.
-Buffers (arrays) in \verb|pyopencl| do not retain type information, so the $\cloquence$ extension adds type information, mirroring the API of \verb|numpy| arrays.
-It also hides many of the most verbose components of the OpenCL interface by providing reasonable defaults, such as an implicit global context and a default queue associated with each context.
-
-Using these simplifications, generic $\cloquence$ OpenCL functions like \verb|map| can be called directly, without specialization as in the previous section.
-Figure \ref{py} shows a full OpenCL program written using these bindings.
-By way of comparison, the same program written using OpenCL directly is two orders of magnitude larger and correspondingly more complex.
-Not shown are several additional conveniences, such as delegated kernel sizing and In and Out constructs that can reduce the size and improve the clarity of this code further -- due to space constraints, the reader is referred to the language documentation for additional details.
-
-\begin{figure}
-\small{\begin{verbatim}
-from figure1 import map
+Developers must annotate functions that meet the requirements to be called from a host with the \verb|__kernel| attribute.
+The \verb|Ace.OpenCL.OpenCL| backend is able to check these requirements and add this annotation automatically.
+Additionally, several types (notably, \verb|double|) and specialized functions require that an OpenCL extension be enabled with a \verb|#pragma|.
+The OpenCL backend can also  automatically infer many of these cases as well and add the appropriate \verb|#pragma| declaration, consistent with the principle of conciseness followed for the Ace design as a whole.
+An example of this can be seen in Listing \ref{mapout}.
+
+\subsection{Metaprogramming in Ace}
+\begin{codelisting}
+\begin{lstlisting}[commentstyle=\color{mauve}]
+from ace.OpenCL import OpenCL
+import ast, ace.astx as astx
+
+plus = OpenCL.fn.from_string("""
+def plus(a, b):
+    return a + b
+""")
+
+add5_ast = astx.specialize(plus.ast, b=5)
+add5 = OpenCL.fn.from_ast(add5_ast)
+\end{lstlisting}
+\caption{Metaprogramming with Ace, showing how to construct generic functions from strings and abstract syntax trees, and how to manipulate syntax trees at compile time.}
+\label{metaprogramming}
+\end{codelisting}
+Metaprogramming refers to the practice of writing programs that manipulate other programs.
+There are a number of use cases for this technique, including domain-specific optimizations and code generation for programs with a repetitive structure that cannot easily be captured using available abstractions.
+OpenCL itself in particular relies on code generation as a fundamental mechanism, partially justifying its lack of support for higher-order programming \cite{pyopencl}.
+Ace supports programmatic compilation and higher-order constructs, as described above, and language extension, which we describe below, so several use cases for this feature have been eliminated.
+However, we demonstrate cases where this form of metaprogramming is useful in the case study described in Section \ref{casestudy}.
+
+An Ace function can be constructed from a string containing its source using the \verb|from_source| variant of \verb|fn|, as demonstrated in Listing \ref{metaprogramming} on Lines 4-7.
+It can also be constructed directly from a Python abstract syntax tree (AST), available via the Python standard \verb|ast| package, using the \verb|from_ast| variant of \verb|fn|, demonstrated on Line 10.
+The AST here is generated programmatically by transforming the syntax tree of the \verb|plus| function such that the argument \verb|b| becomes a constant, \verb|5|.
+This transformation, called \verb|specialize|, as well as several others are distributed with Ace in the \verb|ace.astx| module for convenience.
+\subsection{Direct Invocation from Python}\label{direct}
+
+\begin{codelisting}
+\begin{lstlisting}
 import numpy as np
-import clq.opencl.pyopencl as cl
+import ace.OpenCL.bindings as cl
+from listing3 import map
+from listing4 import add5 # or Listing 7
 
 ctx = cl.Context.for_device(0, 0)
-d_in = ctx.to_device(np.ones(1024))
-d_out = ctx.alloc(like=d_x)
-map(d_in, d_out, clq.opencl.sin, 
+d_in = ctx.to_device(np.ones((1024,)))
+d_out = ctx.alloc(like=d_in)
+map(d_in, d_out, add5, 
     global_size=d_in.shape, local_size=(128,))
 out = ctx.from_device(d_out)
-\end{verbatim}}
-\caption{A full OpenCL program using the $\cloquence$ Python bindings, including data transfer and kernel invocation.}
+\end{lstlisting}
+\caption{A full OpenCL program using the \texttt{Ace.OpenCL} Python bindings, including data transfer and  invocation of a generic function, \texttt{map}, as a kernel without compiling it first.}
 \label{py}
-\end{figure}
-
-\subsection{Code Generation in $\cloquence$}
-Metaprogramming refers to the practice of writing programs that manipulate other programs.
-There are a number of use cases for this technique, including domain-specific optimizations and code generation for programs with a repetitive structure that cannot easily be packaged using available abstractions.
-OpenCL in particular relies on code generation as a fundamental mechanism, partially justifying the lack of support for higher-order programming.
-
-$\cloquence$ supports code generation from directly within the metalanguage.
-A cl.oquence function can be constructed from a string containing its source using the \verb|clq.from_source| function.
-It can also be constructed directly from a Python abstract syntax tree, available via the standard \verb|ast| package, using the \verb|clq.from_ast| function.
-We discuss a use case for code generation for simulation orchestration in $\cloquence$ in Section 4.
-
-\subsection{Active Language Extensions}
-We now turn our attention to the extension mechanism used by $\cloquence$, as motivated by section \ref{extensions}.
-The $\cloquence$ type checker and compiler interacts directly with active libraries over an interface that gives substantial control over type checking and translation of an operation to a {\it type definition} associated with one of its operands according to a fixed dispatch protocol.
-
-\subsubsection{Dispatch Protocol}
-\begin{figure}\small{
-\begin{verbatim}
-class PtrType(Type):
-  def __init__(self, target_type, addr_space):
-    self.target_type = target_type
+\end{codelisting}
+
+As discussed in the Introduction, a common workflow for professional end-users involves the use of a high-level scripting language for overall workflow orchestration and small-scale data analysis and visualization, paired with a low-level language for performance-critical sections.
+Python is already widely used by professional end-users as the high-level scripting language.
+It has support for executing code written in low-level languages in several ways.
+Developers can call into native libraries using its foreign function interface (FFI) or a library like \verb|pycuda| for code compiled with CUDA, a proprietary language similar to OpenCL targeting nVidia GPUs \cite{pycuda}.
+
+The OpenCL language is designed specifically for this workflow, in that it exposes the compiler directly an API, rather than as a separate executable.
+Developers generate OpenCL source code as strings, compile it programmatically, then execute it using the run-time APIs that OpenCL also provides.
+Python bindings to the OpenCL APIs have been developed in a module called \verb|pyopencl| \cite{pyopencl}.
+
+Ace supports a similar workflow as an alternative to using \verb|acecc| as described above.
+Both generic functions and concrete functions written for a backend that supports the direct execution interface (thus far, \verb|OpenCL| and \verb|C99|) can be called like regular Python functions.
+An example of this for the generic \verb|map| function defined in Listing \ref{map} is shown in Listing \ref{py}, with the call itself on Lines 9-10.
+The first two arguments to \verb|map| are OpenCL buffers, generated using a simplified wrapper to the \verb|pyopencl| APIs on Lines 7-8.
+This wrapper associated type information with each buffer, and this information is used to implicitly compile \verb|map| as appropriate.
+That is, explicit calls to the \verb|compile| method, as we have been showing, are unnecessary when using this method of invocation.
+The final two keyword arguments on Line 10 are parameters that OpenCL requires for execution that determine the number of threads (called the {\em global size}) and the grouping of these threads (the {\em local size}).
+
+By way of comparison, the same program written using the OpenCL C API directly is an orders of magnitude larger and correspondingly more complex.
+A full implementation of the logic of \verb|map| written using the \verb|pyopencl| bindings and metaprogramming techniques as described in \cite{pyopencl} is twice as large and significantly more difficult to comprehend than the code we have shown thus far.
+Not shown are several additional conveniences, such as delegated kernel sizing and \verb|In| and \verb|Out| constructs that can reduce the size and improve the clarity of this code further -- the reader is referred to the language documentation for additional details.
+
+\section{Ace for Researchers}
+Thus far, we have been largely discussing the OpenCL module in our examples.
+However, Ace gives no preferential treatment to this module; it is implemented entirely using the user-facing mechanisms described in this section.
+
+Most programming languages are {\em monolithic} -- the set of available primitives is determined by the language designers, and users must combine these primitives to produce desired behaviors.
+Although a number of very flexible primitives have been developed (e.g.
+object systems), these have proven insufficient in specialized domains where developers and researchers need fine control over how certain operations are type checked and translated.
+High-performance computing is an example of such a field, since both correctness and performance have been difficult to achieve in general, and are topics of active research.
+As discussed in the Introduction, the proliferation of parallel programming languages, rather than library-based abstractions, indicates that researchers often find the abstractions in existing languages insufficient for their needs.
+
+In view of these issues, Ace is fundamentally {\em extensible}, rather than monolithic.
+Users can introduce new primitive types and operations, fully controlling how they are type checked and translated.
+The target of translation can also be specified modularly by the user.
+Because Ace libraries can contain compile-time logic, written in the metalanguage as described in the previous section, these primitive definitions can be distributed within normal modules.
+
+\subsection{Active Typechecking and Translation (\ATT)}
+
+When the compiler encounters an expression, it must first verify that it typechecks, then translate it to produce code in a target language.
+Rather than containing fixed logic for this, however, the Ace compiler defers this responsibility to the {\it type} of a subexpression whenever possible, according to a fixed {\em dispatch protocol} for each syntactic form.
+Below are examples of the rules that comprise the Ace dispatch protocol.
+Due to space constraints, we do not list the entire dispatch protocol, which contains a rule for each possible syntactic form.
+\begin{itemize}
+\item Responsibility over a {\bf unary operation} like \verb|-x| is handed to the type assigned to the operand, \verb|x|.
+\item Responsibility over {\bf binary operations} is first handed to the type assigned to the left operand.
+If it indicates that it does not understand the operation, the type assigned to the right operand is handed responsibility, with a different method call\footnote{Note that this operates similarly to the Python run-time operator overloading protocol; see Related Work.}.
+\item Responsibility over {\bf attribute access} (\texttt{obj.attr}) and {\bf subscript access}, (\texttt{obj[idx]}) is handed to the type assigned to \texttt{obj}.
+\end{itemize}
+
+\subsubsection{Active Typechecking}
+During the typechecking phase, the type of the primary operand, as determined by the dispatch protocol, is responsible for assigning a type to the expression as a whole.
+Let us consider the \verb|map| function from Listing \ref{map} once again.
+When it is compiled on Line 9 of Listing \ref{mapadd5dbl}, its first two argument types are given as \verb|global_ptr(double)|.
+As described in Section \ref{types}, this type, abbreviated \verb|A|, is an instance of \verb|ace.OpenCL.GlobalPtrType| which inherits from \verb|ace.OpenCL.PtrType| and ultimately from \verb|ace.Type|.
+So when the compiler encounters the expression \verb|input[gid]| on Line 6, it follows the dispatch protocol just described and assigns responsibility over typechecking it to \verb|A|.
+This is done by calling the \verb|resolve_|$X$ method of the responsible type, where $X$ is the syntactic form of the expression.
+In this case, the expression is of the \verb|Subscript| form, so the compiler calls \verb|A.resolve_Subscript|.
+
+\begin{codelisting}
+\begin{lstlisting}
+import ace, ace.astx as astx
+
+class PtrType(ace.Type):
+  def __init__(self, T, addr_space):
+    self.target_type = T
     self.addr_space = addr_space
 
-  def verify_Subscript(self, context, node):
-    slice_type = context.validate(node.slice)
+  def resolve_Subscript(self, context, node):
+    slice_type = context.resolve(node.slice)
     if isinstance(slice_type, IntegerType):
@@
     else:
-      raise TypeError('<error message here>')
+      raise TypeError('<error message>', node)
 
@@
     slice = context.translate(node.slice)
-    return copy_node(node, 
-      value = value,
-      slice = slice,    
-      code = value.code + '['+slice.code+']')
-\end{verbatim}}
-\caption{A portion of the implementation of OpenCL pointer types implementing subscripting logic using the $\cloquence$ extension mechanism.}
+    return astx.copy_node(node, 
+      value=value, slice=slice,    
+      code=value.code + '[' + slice.code + ']')
+
+  # ...
+
+class GlobalPtrType(PtrType):
+  def __init__(self, T):
+    PtrType.__init__(self, T, '__global')
+\end{lstlisting}
+\caption{A portion of the implementation of OpenCL pointer types implementing subscripting logic using the Ace extension mechanism.}
 \label{pointers}
-\end{figure}
-
-When the compiler encounters an expression, it must first verify that it type checks, then generate code for it.
-Rather than containing fixed logic for this, however, the compiler defers this responsibility to the {\it type} of a subexpression whenever possible.
-Below are example from the $\cloquence$ dispatch protocol.
-Due to space constraints, we do not list the entire dispatch protocol.
-\begin{itemize}
-\item Responsibility over a {\bf unary operation} like \verb|-x| is handed to the type assigned to the operand, \verb|x|.
-\item Responsibility over {\bf binary operations} is first handed to the type assigned to the left operand.
-If it indicates that it does not understand the operation, the type assigned to the right operand is handed responsibility, with a different method call\footnote{Note that this operates similarly to Python's operator overloading protocol.}.
-\item Responsibility over {\bf attribute access}, \texttt{obj.attr}, and {\bf subscript access}, \texttt{obj[idx]}, is handed to the type assigned to \texttt{obj}.
-\end{itemize}
-
-\subsubsection{Verification and Translation}
-A type in $\cloquence$ is a metalanguage {\it instance} of a Python class deriving from \verb|clq.Type|.
-The compiler hands control over an expression to a type by calling a method corresponding to the syntactic form of the expression according to the above dispatch protocol.
-Figure \ref{pointers} gives the verification and translation logic governing subscript access for pointer types in OpenCL.
-
-During the verification phase, the type of the primary operand, as determined by the dispatch protocol, is responsible for assigning a type to the expression as a whole.
-In this case, it recursively assigns a type to the slice operand.
-If it is assigned an integer type, the pointer's target type is assigned to the expression as a whole.
-Otherwise, a type error is raised with the provided error message (an ability that can be used by extension authors to satisfy the error message criteria described in Section \ref{errors}.)
-
-If verification succeeds, the compiler must subsequently translate the $\cloquence$ expression into an expression in the output language, here OpenCL.
-It does so by again applying the dispatch protocol to call a method prefixed with \verb|translate_|.
+\end{codelisting}
+
+The relevant portion of \verb|ace.OpenCL.GlobalPtrType| is shown in Listing \ref{pointers}.
+The \verb|verify_Subscript| method on Line 8 receives a context and the syntax tree of the node itself as input.
+The context contains information about other variables in scope, as well as other potentially relevant information, and also contains a method, \verb|resolve_type|, that can be used recursively resolve the types of subexpressions.
+On Line 9, this method is used to resolve the type of the slice subexpression, \verb|gid|, which is the machine-dependent integer type \verb|size_t| as discussed in Section II.E.
+On Line 10, it confirms that this type is an instance of an integer type.
+Thus, it assigns the whole expression, \verb|input[gid]|, the target type of the pointer, \verb|double|.
+Had a user attempted to index \verb|input| using a non-integer value, the method would take the other branch of the conditional and raise a type error with a relevant user-defined error message on Line 13.
+
+\subsubsection{Active Translation}
+Once typechecking a method is complete, the compiler must subsequently translate each Ace source expression into an expression in the target language, OpenCL in the examples thus far.
+It does so by again applying the dispatch protocol and calling a method of the form \verb|translate_|$X$, where $X$ is the syntactic form of the expression.
 This method is responsible for returning a copy of the expression's ast node with an additional attribute, \verb|code|, containing the source code of the translation.
-In this case, it is simply a direct translation to the corresponding OpenCL attribute access, using the recursively-determined translations of the operands.
-More sophisticated abstractions may insert arbitrarily complex statements and expressions.
+In this case, it is simply a direct translation to the corresponding OpenCL attribute access (Line 20), using the recursively-determined translations of the operands (Lines 16-17).
+More sophisticated abstractions may insert arbitrarily complex statements and expressions during this phase.
 The context also provides some support for non-local effects, such as new top-level declarations (not shown.)
 
-\subsubsection{Modular Backends}\label{backends}
-Thus far, we have discussed using OpenCL as a backend with $\cloquence$.
+\subsection{Active Backends}\label{backends}
+Thus far, we have discussed using OpenCL as a backend with Ace.
 The OpenCL extension is the most mature as of this writing.
-However, $\cloquence$ supports the introduction of new backends in a manner similar to the introduction of new types, by extending the \verb|clq.Backend| base class.
+However, Ace supports the introduction of new backends in a manner similar to the introduction of new types, by extending the \verb|clq.Backend| base class.
 Backends are provided as the first argument to the \verb|@clq.fn| decorator, as can be seen in Figure \ref{map}.
@@
 
-\subsubsection{Use Cases}
+\subsection{Use Cases}
 The development of the full OpenCL language using only the extension mechanisms described above provides evidence of the power of this approach.
@@
 Many of these are currently available either via inconvenient libraries or in standalone languages.
-With the $\cloquence$ extension mechanism, we hope to achieve robust, natural implementations of many of these mechanisms within the same language.
+With the Ace extension mechanism, we hope to achieve robust, natural implementations of many of these mechanisms within the same language.
 
@@
 These languages provide first-class support for accessing data transparently across a massively parallel cluster, which is verbose and poorly supported by standard C.
-The extension mechanism of $\cloquence$ allows inelegant library-based approaches such as the Global Arrays library to be hidden behind natural wrappers that can use compile-time information to optimize performance and verify correctness.
+The extension mechanism of Ace allows inelegant library-based approaches such as the Global Arrays library to be hidden behind natural wrappers that can use compile-time information to optimize performance and verify correctness.
 We have developed a prototype of this approach using the C backend and hope to expand upon it in future work.
@@
 \paragraph{Domain-Specific Type Systems}
-$\cloquence$ is a statically-typed language, so a number of domain-specific abstractions that promise to improve verifiability using types, as discussed in Section \ref{verifiability}, can be implemented using the extension mechanism.
+Ace is a statically-typed language, so a number of domain-specific abstractions that promise to improve verifiability using types, as discussed in Section \ref{verifiability}, can be implemented using the extension mechanism.
 We hope that this will allow advances from the functional programming community to make their way into the professional end-user community more quickly, particularly those focused on scientific domains (e.g.
@@
 Existing compilers make implementing and distribution such optimizations difficult.
-With active libraries in $\cloquence$, optimizations can be distributed directly with the libraries that they work with.
+With active libraries in Ace, optimizations can be distributed directly with the libraries that they work with.
 For instance, we have implemented substantial portions of the NVidia GPU-specific optimizations described in \cite{yang2010gpgpu} as a library that uses the extension mechanism to track affine transformations of the thread index used to access arrays, in order to construct a summary of the memory access patterns of the kernel, which can be used both for single-kernel optimization (as in \cite{yang2010gpgpu}) and for future research on cross-kernel fusion and other optimizations.
@@
 
-\subsection{Availability}
-$\cloquence$ is available under the LGPL license and is developed openly and collaboratively using the popular Github platform\footnote{https://github.com/cyrus-/cl.oquence}.
-Documentation and other learning material is being made available at http://cl.oquence.org/.
-
 \section{Case Study: Neurobiological Circuit Simulation}
 An important criteria that practitioners use to evaluate a language or abstraction, as discussed in Section \ref{social}, is whether significant case studies have been conducted with it.
-In this section, we briefly (due to space limitations) discuss an application of the $\cloquence$ OpenCL library, Python host bindings and code generation features for developing a modular, high-performance scientific simulation library used to simulate  thousands of parallel realizations of a spiking neurobiological circuit on a GPU.
+In this section, we briefly (due to space limitations) discuss an application of the Ace OpenCL library, Python host bindings and code generation features for developing a modular, high-performance scientific simulation library used to simulate  thousands of parallel realizations of a spiking neurobiological circuit on a GPU.
 
@@
 \subsection{The {\sf cl.egans} Simulation Library}
-In order to eliminate these issues while retaining the performance profile of the GPU-accelerated code, the project was ported to $\cloquence$.
+In order to eliminate these issues while retaining the performance profile of the GPU-accelerated code, the project was ported to Ace.
 Rather than using preprocessor directives to control the code contained in the final GPU kernels used to execute the simulation and data analyses, the group was able to develop a more modular  library called {\sf cl.egans}\footnote{...after {\it c.
@@
 
-Implementations of these classes do not evaluate the simulation logic directly, but rather contain methods that generate $\cloquence$ source code for insertion at various points, called {\it hooks}, in the final simulation kernel.
+Implementations of these classes do not evaluate the simulation logic directly, but rather contain methods that generate Ace source code for insertion at various points, called {\it hooks}, in the final simulation kernel.
 The hook that code is inserted into is determined by the method name, and code can be inserted into any hook defined anywhere upstream in the simulation tree.
@@
 
-Using this $\cloquence$-based framework, the benefits of the Brian-based workflow were recovered without the  corresponding decrease in performance relative to the previous CUDA-based solution, leading ultimately to a satisfying solution for the group conducting this research.
-
-\begin{figure}
-\small{
-\begin{verbatim}
+Using this Ace-based framework, the benefits of the Brian-based workflow were recovered without the  corresponding decrease in performance relative to the previous CUDA-based solution, leading ultimately to a satisfying solution for the group conducting this research.
+
+\begin{codelisting}
+\begin{lstlisting}
 # Create the root node of the simulation
@@
 probe = StateVariableProbeCopyback(e_synapse)
-\end{verbatim}}
+\end{lstlisting}
 \caption{An example of a nested simulation tree, showing that specifying a simulation is both simple and modular.}
 \label{spec}
-\end{figure}
-
-\begin{figure}\small{
-\begin{verbatim}
+\end{codelisting}
+
+\begin{codelisting}
+\begin{lstlisting}
 class SpikingModel(Model):
@@
   def in_model_code(self, g):
-    """
-    idx_state = idx_model + (realization_num - 
-        realization_start)*count
-    """ << g  # << appends to code generator, g
-    self.insert_cg_hook("read_incoming", g)
-    self.insert_cg_hook("read_state", g)
-    self.insert_cg_hook("calculate_inputs", g)
-    self.insert_cg_hook("state_calculations", g)
-    self.insert_cg_hook("spike_processing", g)
+    "idx_state = idx_model + count*" << g
+    "(realization_n - realization_start)" << g
+    self.insert_hook("read_incoming", g)
+    self.insert_hook("read_state", g)
+    self.insert_hook("calculate_inputs", g)
+    self.insert_hook("state_calculations", g)
+    self.insert_hook("spike_processing", g)
   # ...
-\end{verbatim}}
+\end{lstlisting}
 \caption{An example of a hook that inserts code and also inserts new, nested hooks for downstream simulation nodes  below that.}
 \label{impl}
-\end{figure}
+\end{codelisting}
 
 \section{Related Work}
+\subsection{Active Libraries in Ace}
+Libraries that have such capabilities has been called {\it active libraries} in prior proposals \cite{activelibraries}.
+A number of  projects, such as Blitz++, have taken advantage of the C++ preprocessor and template-based metaprogramming system to implement domain-specific optimizations.
+In Ace, we replace these brittle mini-languages with a general-purpose language.
+This allows for several interesting uses that we discuss in the following sections.
+
+\subsubsection{Structural Polymorphism}
+In Section \ref{hof}, we discussed several strategies for achieving {\it polymorphism} -- the ability to create functions and data structures that operate over more than a single type.
+In Ace, all functions are implicitly polymorphic and can be called with arguments of {\it any type that supports the operations used by the function}.
+For example, in Figure \ref{map}, \verb|in| can be any type that supports indexing by a variable of type \verb|size_t| to produce a value of a type that can be passed into \verb|fn|, which must then produce a value consistent with indexing into \verb|out|.
+OpenCL pointer types are consistent with these constraints, for example.
+Although powerful, this also demonstrates a caveat of this approach -- that it is more difficult to give a function a concise signature, because arguments are constrained by capability, rather than to a single type \cite{malayeri2009structural}.
+
+Structural typing can be compared to the approach taken by dynamically-typed languages that rely on ``duck typing''.
+It is more flexible than the parametric polymorphism found in many functional languages and in languages like Java (which only allow polymorphic functions that are valid for {\it all} possible types), but is of comparable strength to the template system found in C++.
+It can be helpful to think of each function as being preceded by an implicit template header that assigns each argument its own unique type parameter.
+At function call sites, these parameters are implicitly specialized with the types of the provided arguments.
+This choice is again motivated by the criteria of conciseness given in Section \ref{syntax}.
+
+\subsection{Type-Level Computation} %Haskell, Ur and $\Omega$mega
+System XX with simple case analysis provides the basis of type-level computation in Haskell (where type-level functions are called type families \cite{Chakravarty:2005:ATC}).
+Ur uses type-level records and names to support typesafe metaprogramming, with applications to web programming \cite{conf/pldi/Chlipala10}.
+$\Omega$mega adds algebraic data types at the type-level, using these to increase the expressive power of algebraic data types at the expression level \cite{conf/cefp/SheardL07}.
+Dependently-typed languages blur the traditional phase separation between types and expressions, so type-level computation is often implicitly used (though not always in its most general form, e.g.
+Deputy \cite{conf/icfp/ChenX05}, ATS \cite{conf/esop/ConditHAGN07}.)
+
+\subsection{Run-Time Indirection}
+{\it Operator overloading} \cite{vanWijngaarden:Mailloux:Peck:Koster:Sintzoff:Lindsey:Meertens:Fisker:acta:1975} and {\it metaobject dispatch} \cite{Kiczales91} are run-time protocols that translate operator invocations into function calls.
+The function is typically selected according to the type or value of one or more operands.
+These protocols share the notion of {\it inversion of control} with type-level specification.
+However, type-level specification is a {\it compile-time} protocol focused on enabling specialized verification and implementation strategies, rather than simply enabling run-time indirection.
+
+\subsection{Term Rewriting Systems}
+Many languages and tools allow developers to rewrite expressions according to custom rules.
+These can broadly be classified as {\it term rewriting systems}.
+Macro systems, such as those characteristic of the LISP family of languages \cite{mccarthy1978history}, are the most prominent example.
+Some compile-time metaprogramming systems also allow users to manipulate syntax trees (e.g.
+MetaML \cite{Sheard:1999:UMS}), and external rewrite systems also exist for many languages.
+These facilities differ from type-level specification in one or more of the following ways:
+
+\begin{enumerate}
+\item In type-level specification, the type of a value is determined separately from its representation; in fact, the same representation may be generated by multiple types. 
+\item We draw a distinction between the metalanguage, used to specify types and compile-time logic, the source grammar, used to describe run-time behavior, and the internal language, used to implement this behavior.
+Term rewriting systems generally do not draw this distinction.
+By doing so, each component language can be structured and constrained as appropriate for its distinct role, as we show.
+%\item With type-level specification, dispatch to a type-level function occurs implicitly on the basis of the structure of an expression.
+In contrast, most term-rewriting systems operate by  explicit invocation of a macro or specialized syntax.
+Some LISP macro systems have explored pattern-based dispatch (e.g.
+A*\cite{fowler2010domain}, EPP\cite{fowler2010domain}) and macro systems for object-oriented languages, like OpenC++ \cite{fowler2010domain} and OpenJava \cite{fowler2010domain}, do offer a somewhat limited form of operation-based dispatch.
+\item Many common macro systems and metaprogramming facilities operate at run-time.
+Compilers for some forms of LISP employ aggressive compile-time specialization techniques to attempt to minimize this overhead.
+Static and staged term-rewriting systems also exist (e.g.
+OpenJava\cite{TatM:OpenJCBMSJ}, Template Haskell\cite{SheardPeytonJones:Haskell-02}, MetaML \cite{Sheard:1999:UMS} and others).
+\end{enumerate}
+
+\subsection{Language Frameworks}
+When the mechanisms available in an existing language prove insufficient, researchers and domain experts must design a new language.
+A number of tools have been developed to assist with this task, including compiler generators, language workbenches and domain-specific language frameworks (cf \cite{fowler2010domain}).
+
+A major barrier to adoption is the fact that interoperability is intrinsically problematic.
+Even languages which target a common platform, such as the Java Virtual Machine, can only interact using its limited set of primitives.
+Specialized typing rules are not checked at language boundaries, performance often suffers, and the syntax can be unnatural, particularly for languages which differ significantly from the platform's native language (e.g.
+Java).
+
+Instead of focusing on defining standalone languages, type-level specification gives greater responsibility in a granular manner to libraries.
+In this way, a range of constructs can coexist within the same program and, assuming that it can be shown by some method that various constructs are safely composable, be mixed and matched.
+The main limitation is that the protocol requires defining a fixed source grammar, whereas a specialized language has considerable flexibility in that regard.
+Nevertheless, as Ace shows, a simple grammar can be used quite flexibly.
+\subsection{Extensible Compilers}
+An alternative methodology is to implement language features granularly as compiler extensions.
+As discussed in Section 1, existing designs suffer from the same problems related to composability, modularity\-, safety and security as extensible languages, while also adding the issue of language fragmentation.
+
+Type-level specification can in fact be implemented within a compiler, rather than provided as a core language feature.
+This would resolve some of the issues, as described in this paper.
+However, by leveraging type-level computation to integrate the protocol directly into the language, we benefit from common module systems and other shared infrastructure.
+We also avoid the fragmentation issue.
+\subsection{Specification Languages}
+Several {\it specification languages} (or {\it logical frameworks}) based on these theoretical formulations exist, including the OBJ family of languages (e.g.
+CafeOBJ \cite{Diaconescu-Futatsugi01}).
+They provide support for verifying a program against a language specification, and can automatically execute these programs as well in some cases.
+The  language itself specifies which verification and execution strategies are used.
+
+Type-determined compilation takes a more concrete approach to the problem, focusing on combining {\it implementations} of different\- logics, rather than simply their specifications.
+In other words, it focuses on combining {\it type checkers} and {\it implementation strategies} rather than more abstract representations of a language's type system and dynamic semantics.
+In Section 4, we outlined a preliminary approach based on proof assistant available for the type-level language to unify these approaches, and we hope to continue this line of research in future work.
 
 \section{Conclusion}
+In addition to the novel architecture of Ace as a whole, we note several individually novel features introduced here: %in this paper:
+
+\begin{itemize}
+\item The \ATT~mechanism, which is a generalization of the concept of active libraries \cite{activelibraries}  where types are metalanguage objects.
+\item The method Ace uses to eliminate the need for type annotations in most cases, which combines a form of type inference with type propagation.
+\item The method Ace uses check correctness of generated code by checking representational consistency constraints associated with types, detailed in Section \ref{repcon}. 
+\item The type-aware simulation orchestration techniques used in the \texttt{cl\_egans} library, described in Section \ref{clegans}.
+\end{itemize}
+
+Readers familiar with the Python programming language will recognize the style of syntax used in Figure \ref{map}.
+In fact, Ace uses the Python grammar and parsing facilities directly.
+Several factors motivated this design decision.
+First, Python's syntax is widely credited as being particularly simple and readable, due to its use of significant whitespace and conventional mathematical notation.
+Python is one of the most widely-used languages in scientific computing, so its syntax is already familiar to much of the field.
+And significantly, a large ecosystem of tools already exist that work with Python files, such as code editors, syntax highlighters, style checkers and documentation generators.
+These can be used without modification to work with Ace files.
+Therefore, by re-using an existing, widely-used grammar, we are able to satisfy many of the design criteria described in Section \ref{syntax} and the adoption criteria described in Section \ref{tools} without significant development effort.
+
 Professional end-users demand much from new languages and abstractions.
 In this paper, we began by generating a concrete, detailed set of design and adoption criteria that we hope will be of broad interest and utility to the research community.
-Based on these constraints, we designed a new language, $\cloquence$, making several pragmatic design decisions and utilizing advanced techniques, including type inference, structural typing, compile-time metaprogramming and active libraries, to uniquely satisfy many of the criteria we discuss, particularly those related to extensibility.
+Based on these constraints, we designed a new language, Ace, making several pragmatic design decisions and utilizing advanced techniques, including type inference, structural typing, compile-time metaprogramming and active libraries, to uniquely satisfy many of the criteria we discuss, particularly those related to extensibility.
 We validated the extension mechanism with a mature implementation of  the entirety of the OpenCL type system, as well as preliminary implementations of some other features.
 Finally, we demonstrated that this language was useful in practice, drastically improving performance without negatively impacting the high-level scientific workflow of a large-scale neurobiological circuit simulation project.
-Going forward, we hope that $\cloquence$ (or simply the key techniques it proposes, by some other vehicle) will be developed further by the community to strengthen the foundations upon which new abstractions are implemented and deployed into professional end-user development communities.
+Going forward, we hope that Ace (or simply the key techniques it proposes, by some other vehicle) will be developed further by the community to strengthen the foundations upon which new abstractions are implemented and deployed into professional end-user development communities.
+
+\section{Availability}
+Ace is available under the LGPL license and is developed openly and collaboratively using the popular Github platform at \url{https://github.com/cyrus-/ace}.
+Documentation, examples and other learning materials will be available at \url{http://acelang.org/}.
+{\color{red}(by the time of the conference)}
+
+\section{Acknowledgments}
+CO was funded by the DOE Computational Science Graduate Fellowship under grant number DE-FG02-97ER25308. NF was funded by ...
 

### 2012-10-14 ace/ace.tex
ace: Revised through section 2.C

@@
 We introduce the Ace compilation environment and demonstrate its suitability as a foundational tool for both research and practice in high-performance computing.
-Ace consists of a statically-typed programming language with {\em user-extensible semantics} defined using a compile-time meta\-language, Python.
+Ace consists of a statically-typed programming language with {\em user-extensible semantics} specified via a compile-time meta\-language, Python.
 Users specify new primitive types and their operations by equipping type definitions, which are first-class objects in the metalanguage, with  methods that the compiler selectively invokes when type checking and translating expressions, a mechanism we call {\em active type-checking and translation (\ATT)}.
-We demonstrate the flexibility of this mechanism by implementing primitives from several widely-known languages, including the entirety of OpenCL.
-Ace also supports more general forms of metaprogramming, and Ace functions can be compiled and launched directly from Python with standard numeric data structures as arguments.
-Using each of these features, we designed a scientific simulation framework that allows users to modularly specify, compile and orchestrate the execution of parameterized families of scientific simulations on clusters of GPUs.
+We demonstrate the flexibility of this mechanism by implementing primitives from several widely-known languages, including the entirety of OpenCL, as libraries.
+Ace also supports more general forms of metaprogramming, and functions can be launched directly from Python with standard numeric data structures as arguments.
+Using these features, we designed a scientific simulation framework that allows users to modularly specify, compile and orchestrate the execution of parameterized families of scientific simulations on clusters of GPUs.
 This framework has been used to successfully conduct large-scale, high-performance neuroscience simulations, providing initial evidence that Ace is useful in practice today.
@@
 
-Computer-aided simulation and data analysis techniques have transformed science and engineering.
+Computer-aided simulations and data analysis techniques have transformed science and engineering.
 Surveys show that scientists and engineers now spend up to 40\% of their time writing software \cite{howison2011scientific, hannay2009scientists}.
@@
 
-The research community has proposed dozens of novel language features and parallel programming abstractions that aim to  more equitably balance important concerns related to performance, portability, verifiability and ease-of-use.
-Unfortunately, professional end-users are hesitant to adopt these, with many generally skeptical regarding the practicality of new approaches that originate in the research community.
+The research community has proposed dozens of novel language features and parallel programming abstractions that aim to  more equitably balance important concerns relating to performance, portability, verifiability and ease-of-use.
+Unfortunately, professional end-users are hesitant to adopt these, with many being generally skeptical regarding the practicality of new approaches that come from the research community.
 This viewpoint was perhaps most succinctly expressed by a participant in a recent study by Basili et al.
 \cite{basili2008understanding}, who stated ``I hate MPI, I hate C++.
-[But] if I had to choose again, I would probably choose the same.'' Although this sentiment is easy to dismiss as paradoxical, we believe that it demands direct examination by researchers working to advance the practice of high-performance computing.
-
-A major barrier to adoption of new language features and abstractions is that they are often developed and distributed together with a new programming language.
-Some prominent examples of this language-oriented approach include:
+[But] if I had to choose again, I would probably choose the same.'' Although this sentiment is easy to dismiss as paradoxical, we believe that it demands direct examination by researchers working to advance high-performance computing.
+
+Many of the barriers to adoption of new abstractions stem from the fact that new abstractions are often developed and distributed together with a new programming language.
+Some prominent examples where this {\em language-oriented approach} has been followed include:
 \begin{itemize}
-\item UPC \cite{upc}, CAF \cite{caf}, Fortress \cite{fortress}, X10 \cite{x10} and Chapel \cite{chapel}, which support different forms of the partitioned global address space (PGAS) model; 
-\item Actors, CSPs and other forms of message passing are best supported by languages like Erlang, Charm++, Occam and Axum.
-\item Nested data parallelism is best supported by languages like Cilk and NESL.
+\item UPC \cite{upc}, CAF \cite{caf}, X10 \cite{x10} and Chapel \cite{chapel}, built around first-class support for a partitioned global address space (PGAS) programming model; 
+\item Erlang \cite{erlang}, Charm++ \cite{charm}, Occam \cite{occam} and Axum \cite{axum}, built around first-class support for message passing;
+\item Cilk \cite{cilk} and NESL \cite{nesl}, built around first-class support for nested data parallelism; and
+\item Add something about automatic parallelization of functional programs.
 \end{itemize}
 
-Unfortunately, using a specialized research language comes with a host of practical challenges, such as the cost of learning the language, and the lack of well-developed libraries and tools.
-In fact, the most widely adopted parallel programming abstractions have been those that are distributed as libraries, such as MapReduce, Global Arrays and MPI.
-This approach is more practical because it allows developers to progressively introduce parallelism into existing programs, leverage a familiar and well-developed general-purpose programming language when not directly invoking parallel primitives, and utilize different abstractions within a single program without encountering friction inherent to language boundaries.
-Abstractions developed as libraries are limited, however, in that they cannot alter or augment the semantics of the host language itself.
-Thus, abstractions that require new type checking rules and compilation logic for the types and associated operations they introduce are at a relative disadvantage in practice.
-
-Ace aims to solve this problem by allowing developers to define and distribute new first-class language constructs in a safely composable manner directly within user libraries using a mechanism called {\em active typechecking and translation (\ATT)}.
-In particular, \ATT~allows library authors to extend the semantics of the  language while retaining a fixed but flexible grammar.
-Aided by a form of type inference, the Ace grammar -- a large subset of Python's -- is flexible enough that we are able to define all of the primitives of the C99 and OpenCL languages in a natural manner.
-These primitives can then be supplemented with higher-level user-defined primitives, examples of which we discuss in Section \ref{abstractions}, as desired.
-Notably, these primitives are imported via the normal library loading mechanism; they are not defined within domain-specific languages, nor as extensions for a particular compiler for a language.
-
-%To enable this and other, Ace programs contain both compile-time and run-time program logic.
-The compile-time logic, written in the {\em metalanguage}, is the top-level code in each file.
-Types are first-class objects in the metalanguage, and the logic needed to typecheck and compile the operators associated with a type is bundled with the type definition itself.
-Abstractions can target existing backend languages, using its existing compilation infrastructure for low-level optimizations, so only the high-level semantics need to be encoded by researchers.
-Consistent with its design intent, we demonstrate the use of OpenCL as a flexible backend language and compiler framework \cite{opencl}.
-%Here, users can define and import new primitive types and operators, as well as do other types of metaprogramming tasks like generate functions programmatically, perform arbitrary code transformations.
-% Ace also permits an interactive mode of use whereby the functions can also be executed immediately after being defined, a common usage scenario for professional end-users.
-
-Although the \ATT~mechanism, a key feature of Ace, is designed for use by researchers and domain experts, Ace is designed to be highly practical for professional end-user developers as well.
+There is evidence suggesting that developers would prefer first-class support for a parallel abstraction over a library-based solution, if all else is held constant \cite{ppfirstclass}.
+Unfortunately, adopting a specialized research language comes with a host of ancillary challenges.
+Learning a new language can be time-consuming, new languages often lack well-developed libraries and tools, and it can be difficult to port large legacy codebases.
+Indeed, the most widely adopted parallel programming abstractions have been those that are distributed as libraries for an existing, widely-used language, such as MapReduce, Global Arrays and MPI.
+This {\em library-oriented approach} has proven more practical because it allows developers to progressively introduce parallelism into existing programs, leverage a familiar and well-developed general-purpose programming language when not directly invoking parallel primitives, and utilize different abstractions within a single program without encountering the friction inherent to language boundaries.
+
+Ace aims to solve this problem by allowing developers to define and distribute new {\em first-class language constructs}, in a safely composable manner, directly within user libraries, via a mechanism we introduce called {\em active typechecking and translation (\ATT)}.
+In particular, \ATT~is a mechanism that allows developers to  extend the semantics of a language that has a fixed but flexible grammar.
+The Ace grammar is a large subset of Python's  and we show that it is flexible enough that we are able to define all of the primitives of the C99 and OpenCL languages in a natural manner, aided by a form of type inference.
+These primitives can then be supplemented with higher-level user-defined primitives, examples of which we discuss in Section \ref{abstractions}.
+Notably, all of these primitives are imported via the normal library loading mechanism; they are not defined within domain-specific languages, nor as extensions for a particular compiler for a language.
+
+Although the \ATT~mechanism, a key feature of Ace, is designed primarily for use by researchers and domain experts, Ace is intended to be highly practical and ready for adoption by professional end-user developers today.
 It is based on the Python programming language, which enjoys broad adoption in this community.
-Indeed, the language itself is implemented entirely as a Python library, so many existing tools, such as editor modes, documentation tools and style checkers, are available immediately.
-Ace programs can be generated programmatically and called directly from Python, with numeric arrays as arguments, so Ace integrates immediately with existing workflows to speed up performance-critical portions of code that may benefit from static compilation and support for parallelism.
-We demonstrate several of these features by developing a scientific simulation framework, which we use for specifying and orchestrating families of large-scale neurobiological circuit simulations on clusters of GPUs, in Section \ref{clegans}.
-
-%Researchers design and implement novel parallel programming abstractions\- on a regular basis.
-Each abstraction aims to make some class of developers more productive by making appropriate trade-offs between performance, portability, verifiability and ease-of-use.
-It can be observed that the most widely-adopted abstractions to date are those that been implemented as libraries that can be easily deployed within existing, widely-used languages.
-Abstractions that require introducing new primitive constructs, changing the semantics of existing constructs in particular contexts, or controlling compilation in a fine-grained manner cannot take this approach.
-This has led to a proliferation of specialized programming languages, each designed around a few privileged abstractions.
-Unfortunately, such languages have not been widely adopted by practitioners.
-%
-%slash
-%
-%Although researchers often propose higher-level language features and parallel programming abstractions that aim to strike a better balance between raw performance, productivity, code portability and verifiability (see Section 2), end-users remain skeptical of new approaches.
-This viewpoint was perhaps most succinctly expressed by a participant interviewed in a recent study \cite{basili2008understanding}, who stated ``I hate MPI, I hate C++.
-[But] if I had to choose again, I would probably choose the same.'' Although this sentiment is easy to dismiss as paradoxical, we believe that it demands direct examination by those in the research community working to advance the practice of scientific and high-performance computing.
-We begin in Section 2 by developing a set of design and adoption criteria for new languages and abstractions based on prior empirical studies of these {\it professional end-user developers} as well as our observations of characteristics common to successful projects in the past.
-
-%
-% Two types of languages see widespread use: productivity-oriented and performance-oriented
-% Significant diversity in productivity-oriented languages. Valued for clean syntax and high-level abstractions. Usually very slow. Have a FFI to call into a performance-oriented language.
-% Performance-oriented languages -- C or Fortran and their derivatives.
-Direct memory access, low-level abstractions, more verbose.
-Portable (though often need to include platform-specific optimizations, which can lead to messy code).
-Not easily extended with new abstractions and primitives (witness: lots of derivative languages that are not composable because no way to implement these abstractions cleanly)
-% Generally don't support verifiability very well (performance-oriented languages have simple type systems); only exception is functional languages.
-%
-
-% Simplicity (Errors, Syntax, Core Semantics)
-% Extensibility (New Abstractions, Domain-Specific Things, Domain- and Platform-Specific Optimizations)
-% Performance (Need to allow control over data layout and movement, perhaps optionally)
-% Portability (Architectures and operating systems; support new optimizations)
-% Backwards-Compatibility (Existing codebases)
-% Verifiability (type systems, metaprogramming-based, formal semantics, small TCB)
-% Tool Support (syntax highlighting, editors, debuggers, etc.)
-% Social Proof (Will not disappear - open source, significant projects have been developed, community)
-
-%We then introduce a new programming language named Ace (pronounced ``C eloquence'') in Section 3.
-Ace initially targets code paths where low-level languages, particularly OpenCL, would be used today.
-By employing a collection of powerful front-end language and compiler design techniques, making pragmatic design choices, and building on top of established infrastructure, we argue that Ace may uniquely satisfy many of the criteria established in Section 2, particularly those related to ease of use, familiarity, performance and tool support.
-%
-%Unlike many existing languages, Ace is designed to be highly extensible from inception, employing a powerful compile-time language extension and code generation mechanism based around the concept of {\it active libraries} \cite{activelibraries}.
-We argue that this feature may make Ace particularly suitable as a platform for researchers developing new parallel abstractions and other specialized, high-level language features, as well as  development tools.
-%
-%In Section 4 we briefly describe a case study where the language was used to conduct large-scale spiking neural circuit simulations on GPU hardware, demonstrating the feasibility of this approach on a non-trivial problem.
-Finally, conclude in Section 6 with future directions for both the Ace language and the research community.
-%
+Indeed, the language itself is implemented entirely as a Python library, and many existing tools, such as editor modes, documentation tools and style checkers, can be used without modification.
+Ace functions are generated programmatically and can be called directly from Python, with numeric arrays as arguments, so developers can integrate Ace immediately within their existing workflows to define portions of code that may benefit from static compilation and support for parallelism.
+We demonstrate the practicality of Ace in Section \ref{clegans} by describing an Ace-based framework that supports specifying and orchestrating families of large-scale neurobiological circuit simulations on clusters of GPUs.
 
@@
 \subsection{Example 1: Hello, World!}
-To demonstrate the structure of programs and libraries in Ace, we begin in Listing \ref{hello} with a variant of the canonical ``Hello, World!'' example written using the OpenCL module distributed with Ace.
-We note that the OpenCL module, which we use throughout this paper in examples, is simply a library like any other.
+To demonstrate the structure of programs and libraries in Ace, let us begin in Listing \ref{hello} with a variant of the canonical ``Hello, World!'' example written using the OpenCL module distributed with Ace.
+We note that this module, which we use throughout the paper in examples, is simply a library like any other.
 The core of Ace gives no special treatment to it; it is distributed with Ace for convenience.
@@
 Indeed, Ace programs are Python scripts at the top-level, rather than a list of declarations as in most conventional languages.
-In other words, Python is the {\em compile-time metalanguage} of Ace.
-In other words, Ace programs are embedded within Python.
-By making this choice, Ace can  use Python's well-developed module system and associated distribution infrastructure directly (Line 1).
-
-The \verb|main| function defined on Lines 5-7 contains a single call to the \verb|printf| primitive imported from the \verb|ace.OpenCL| module on Line 1.
-The function is introduced using the \verb|@OpenCL.fn| decorator, which indicates that it is a {\em actively-typed} Ace function containing run-time logic written for the \verb|OpenCL| backend, also imported on Line 1.
+In other words, Python is the {\em compile-time metalanguage} of Ace; Ace programs are embedded within Python.
+A consequence of this choice is that Ace can leverage Python's well-developed package system and  distribution infrastructure directly (Line 1).
+
+The \verb|main| function defined on Lines 5-7 contains a single call to the \verb|printf| primitive, imported from the \verb|ace.OpenCL| module on Line 1.
+The function is introduced using the \verb|@OpenCL.fn| decorator, which indicates that it is a statically-typed Ace function containing run-time logic targeting the \verb|OpenCL| backend, also imported on Line 1.
 Without this decorator, the function would simply be a conventional Python function that could be called only at compile-time (doing so would fail here: \verb|printf| is not a Python primitive.)
 
-On Line 8, the {\em generic function} defined on Lines 5-7, is compiled to produce a {\em concrete function} which we also call \verb|main| (overwriting the previous definition).
+On Line 8, this {\em generic function} defined on Lines 5-7, is compiled to produce a {\em concrete function} that we also identify as \verb|main| (overwriting the previous definition).
 The distinction between generic and concrete functions will be explained in our next example.
@@
 \item Produces source code associated with any concrete functions (functions produced using the \verb|compile| method on Line 8 of Listing \ref{hello}) that are in the top-level environment.
-Each backend may produce one or more files (here, only \verb|hello.cl| is produced).
+Each backend may produce one or more files (here, \verb|hello.cl| is produced).
 \end{enumerate}
 
-Use of the \verb|acecc| executable is simply a convenience; the generated code is also available as an attribute of the concrete function: \verb|main.code|.
-We will also show in Section \ref{pyopencl} that, for backend languages with Python bindings, such as OpenCL and C, functions can be executed directly, without an intermediate compilation step, if desired.
+Using the \verb|acecc| executable is simply a convenience; the generated code is also available as an attribute called of the concrete function immediately after it has been defined, accessed as \verb|main.code|.
+We will show in Section \ref{pyopencl} that, for backend languages with Python bindings, such as OpenCL and C, generic functions can be executed directly, without the intermediate compilation step, if desired.
 
@@
 \end{lstlisting}
-\caption{[\texttt{listing3.py}] A generic data-parallel higher-order map function written using the user-defined Ace OpenCL module.}
+\caption{[\texttt{listing3.py}] A generic data-parallel higher-order map function written using OpenCL user module.}
 \label{map}
@@
 
-A, B = global_ptr(double), global_ptr(int)
+A = global_ptr(double); B = global_ptr(int)
 map_add5_dbl = map.compile(A, A, add5.ace_type)
@@
 \end{lstlisting}
-\caption{[\texttt{listing4.py}] The generic map function compiled to map the sin function over global arrays of double-precision floating-point numbers.
-The \texttt{get\_global\_id} and \texttt{sin} functions are OpenCL primitives.}
+\caption{[\texttt{listing4.py}] The generic \texttt{map} function compiled to map the \texttt{add5} function over two  types of input.}
 \label{mapadd5dbl}
@@
 \end{lstlisting}
-\caption{The OpenCL code generated by running \texttt{acecc listing4.py}, or by executing the  \texttt{print listing4.map\_sin\_dbl.code} statement.}
+\caption{[\texttt{listing4.cl}] The OpenCL code generated by running \texttt{acecc listing4.py}.}
 \label{mapout}
@@
 The ``Hello, World!'' example demonstrates the basic structure of Ace programs, but it does not require working with types.
-Listing \ref{map} shows a data-parallel map primitive written using the OpenCL library, introduced above.
-OpenCL is a data-parallel language -- users can define functions, called {\em kernels}, that execute across thousands of threads.
-Each kernel has access to its index, called its {\em global id}, which can be used by the programmer to ensure that each thread operates on different parts of the input data.
-The \verb|map| kernel defined in Listing \ref{map} applies a transfer function, \verb|f|, to the element of the input array, \verb|input| corresponding to its global id, writing the result into the corresponding location in the provided output array, \verb|output|.
-
-As written, \verb|map| is a {\em generic function} (an instance of the class \verb|ace.GenericFn|).
-This means that it does not have types assigned to its arguments.
-Indeed, the functionality of \verb|map| can be defined for many combinations of array types for \verb|input| and \verb|output| and functions.
-In this sense, \verb|map| is actually a family of functions defined for all types assigned to \verb|input|, \verb|output| and \verb|f| for which the operations in the function's body are typesafe.
+Listing \ref{map} shows an imperative, data-parallel map primitive written using the OpenCL library introduced above.
+In OpenCL, users can define functions, called {\em kernels}, that execute across thousands of threads.
+Each kernel has access to a unique index, called its {\em global id}, which can be used by the programmer to ensure that each thread operates on different parts of the input data (Line 5).
+The \verb|map| kernel defined in Listing \ref{map} applies a transfer function, \verb|f|, to the element of the input array, \verb|input|, corresponding to its global id.
+It writes the result of this call into the corresponding location in the provided output array, \verb|output|.
+
+As above, \verb|map| is a {\em generic function} (specifically, an instance of the class \verb|ace.GenericFn|).
+This means that it's arguments have not been assigned types.
+Indeed, the functionality given by the \verb|map| definition is applicable to many combinations of types for \verb|input| and \verb|output| and functions, \verb|f|.
+In this sense, \verb|map| is actually a family of functions defined for all types assignments for \verb|input|, \verb|output| and \verb|f| for which the operations in the function's body are well-defined.
+
 Running \verb|acecc listing3.py| would produce no output, however.
-
 To create a {\em concrete function} (an instance of the class \verb|ace.ConcreteFn|) that can be emitted by the compiler, types must be assigned to each of the arguments of any externally callable functions.
-Listing \ref{mapadd5dbl} shows how to use the \verb|compile| method to specialize \verb|map| to apply the \verb|add5| function, defined on Lines 4-6, to arrays  that reside in global memory (OpenCL has a notion of four different memory spaces).
+Listing \ref{mapadd5dbl} shows how to use the \verb|compile| method to specialize \verb|map| in two different ways to apply the \verb|add5| function, defined on Lines 4-6, to arrays  that reside in global memory (OpenCL has a notion of four different memory spaces).
 Line 9 produces a version specialized for arrays of \verb|double|s and Line 10 produces a version for arrays of \verb|int|s.
@@
 \subsection{Types as Metalanguage Objects}
-The types of the arguments of the functions created on Lines 4.9 and 4.10 were written in pure OpenCL would be \verb|__global double*| and \verb|__global int*| respectively.
-The corresponding types in Ace are \verb|global_ptr(double)| and \verb|global_ptr(int)|, abbreviated as \verb|A| and \verb|B| respectively on Line 8 for convenience (note that here, standard variable assignment in the metalanguage is serving the role that a \verb|typedef| declaration would in a C-like language.)
-
-Types in Ace are values in the metalanguage, Python.
-More specifically, types are {\em object instances} of user-defined classes that inherit from \verb|ace.Type|.
-For example, \verb|global_ptr(double)| is an instance of \verb|ace.OpenCL.GlobalPtrType| instantiated with the target type, \verb|double|, as an argument.
-The type \verb|double| itself (imported from \verb|ace.OpenCL| on Line 4.2) is an instance of \verb|ace.OpenCL.FloatType|.
-As we will describe further in Section \ref{att}, this notion of ``types as metalanguage objects'' is a key to Ace's flexibility.
+The types of the arguments of the functions being compiled on Lines 4.9 and 4.10, if written directly in OpenCL as in Listing \ref{mapout}, are \verb|__global double*| and \verb|__global int*|, respectively.
+The corresponding types in Ace are \verb|global_ptr(double)| and \verb|global_ptr(int)|, abbreviated for convenience as \verb|A| and \verb|B| respectively on Line 8 (note that here, standard variable assignment in the metalanguage is serving the role that a \verb|typedef| declaration would fill in a C-like language.)
+
+\textbf{Types in Ace are values in the metalanguage}, Python.
+More specifically, types are {\em object instances} of user-defined classes that inherit from the Ace-provided \verb|ace.Type| class.
+For example, \verb|global_ptr(double)| is an instance of \verb|ace.OpenCL.GlobalPtrType| instantiated with the target type, \verb|double|, as a constructor argument.
+The types \verb|double| and \verb|int| (imported from \verb|ace.OpenCL| on Line 4.2) are instances of \verb|ace.OpenCL.FloatType| and \verb|ace.OpenCL.IntegerType|, respectively.
+As we will describe further in Section \ref{att}, this notion of types as metalanguage objects is a key to Ace's flexibility.
 

### 2012-10-14 ace/ace.tex
ace: Revised through section 2.G, part of 2.H

@@
   basicstyle=\ttfamily\footnotesize,
-  morekeywords={lambda, self, assert},
+  morekeywords={lambda, self, assert, as},
   numbers=left,
@@
 \subsection{Type Propagation and Higher-Order Functions}
-The type assigned to the third argument, \verb|f|, on Lines 9 and 10, is given as \verb|add5.ace_type|.
+The type assigned to the third argument, \verb|f|, on Lines 4.9 and 4.10, is given as \verb|add5.ace_type|.
 The \verb|ace_type| attribute of a generic function is an instance of \verb|ace.GenericFnType|, the type of Ace generic functions, that corresponds to that particular function, \verb|add5| in this case.
-Generic functions are compiled automatically when they are called.
-That is, when the compiler encounters the call to \verb|f| inside \verb|map| when compiling \verb|map_add5_double|, it compiles a version of \verb|add5| specialized to the \verb|double| type, and similarly with the \verb|int| type when compiling \verb|map_add5_int|.
+Generic functions are compiled automatically when they are called from another function.
+That is, when the compiler encounters the call to \verb|f| inside \verb|map| when compiling \verb|map_add5_double|, it compiles a version of \verb|add5| specialized to the \verb|double| type (seen on Line 5.3), and similarly with the \verb|int| type when compiling \verb|map_add5_int| (on Line 5.16, automatically given a unique name to avoid conflicts).
 This mechanism is called {\em type propagation}.
-In other words, we did not need to run \verb|add5.compile(double)| before compiling \verb|map_add5_dbl| (although doing so would also be valid).
-Only functions that are never called in the process of compiling other functions in a concrete module need type information explicitly provided, whereas \verb|add5| is a function that is only called within \verb|map|.
-
-We note that this scheme allows for a form of higher-order functional programming even when targeting languages, like OpenCL, that have no support for higher-order functions (OpenCL, unlike C99, does not support function pointers.) This works because the \verb|ace.GenericFnType| for one function, such as \verb|add5|, is not the same as the \verb|ace.GenericFnType| for a superficially similar function, such as \verb|add6| (defined as you would expect.) To put it in type theoretic terms, \verb|ace.GenericFnType|s are {\em uniquely inhabited} by a single generic function (and similarly for \verb|ace.ConcreteFnType|s), and thus it is impractical to use them as full first-class values (i.e.
-they cannot put into a run-time array.) In practice, this limitation is rarely a problem, particularly in parallel programming where such forms of specialization are already common in order to avoid the potential performance penalty associated with a function pointer dereference per call.
-
-We note that fully-featured first-class functions can be implemented when targeting a backend that supports them, such as C99, or by using a jump table implementation, but we do not discuss this further due to lack of space.
-%%% TODO: Refer to line numbers in output
+In other words, we did not need to use \verb|add5.compile(double)| before compiling \verb|map_add5_dbl| (although we could if we would like).
+Only functions that are never called in the process of compiling other functions in a module need type information explicitly provided, whereas \verb|add5| is a function that is only called within \verb|map| in Listing \ref{mapadd5dbl}.
+
+We note that this scheme allows for a form of higher-order functional programming even when targeting languages, like OpenCL, that have no support for higher-order functions (OpenCL, unlike C99, does not even support function pointers).
+This works because the \verb|ace.GenericFnType| for one function, such as \verb|add5|, is not the same as the \verb|ace.GenericFnType| for a superficially similar function, such as \verb|add6| (defined as one would expect).
+To put it in type theoretic terms, \verb|ace.GenericFnType|s are singleton types, {\em uniquely inhabited} by a single generic function, and thus it is impractical to use them as full first-class values (i.e.
+they cannot put into a run-time array.) In practice, this is rarely a problem, particularly in parallel programming where explicit specialization is already common in order to avoid the potential performance penalty associated with a function pointer dereference per call.
+We note that fully-featured first-class functions can be implemented when targeting a backend that supports them, such as C99, or simulated by using a jump table implementation in OpenCL, but we do not discuss this further due to lack of space.
 
 \subsection{Type Inference}
-Within the \verb|map| function in Listing \ref{map}, on Line 5, the variable \verb|gid| is initialized with the result of calling the \verb|get_global_id| primitive (OpenCL supports multidimensional global ids, so the argument specifies which dimension to use.)  Note that a type for the \verb|gid| variable is not listed explicitly.
-This is because Ace supports a form of within-function {\em type inference}.
-In this case, \verb|gid| will have type \verb|size_t| because that is the return type of \verb|get_global_id| (as defined in the OpenCL specification, which the \verb|ace.OpenCL| module follows.) This can be seen on Lines 11 and 24 in Listing \ref{mapout}.
-
-Inference is not restricted to single assignments, as in the \verb|map| example.
-Multiple assignments to the same variable with values of differing types and multiple return statements can be unified, such that the variable or return type is given a common supertype.
-For example, in Listing \ref{inference} the variable \verb|y| in the first branch of the conditional is assigned the \verb|int| literal \verb|0|.
+On Line 5 in the generic \verb|map| function in Listing \ref{map}, the variable \verb|gid| is initialized with the result of calling the \verb|get_global_id| primitive (the argument, \verb|0|, is not important for our purposes.)  Note that the type for the \verb|gid| variable is never listed explicitly.
+This is because Ace supports a form of {\em whole-function type inference}.
+In this case, \verb|gid| will be assigned type \verb|size_t| because that is the return type of \verb|get_global_id| (as defined in the OpenCL specification \cite{opencl}, which the \verb|ace.OpenCL| module follows.) The result can be observed on Lines 11 and 24 in Listing \ref{mapout}.
+
+Inference is not restricted within single assignments, as in the \verb|map| example.
+Multiple assignments to the same variable with values of differing types, or multiple return statements, can be unified such that the variable or return type is given a common supertype.
+For example, in the \verb|threshold_scale| function defined in Listing \ref{inference}, the variable \verb|y| in the first branch of the conditional is assigned the \verb|int| literal \verb|0|.
 However, in the second branch of the loop, its type depends on the types of \verb|x| and \verb|scale|.
-We show two examples where this is the case on Lines 10 and 11.
-However, type inference correctly unifies these two types according to the C99 rules governing numeric types (which can be defined by the user, as we show in Section \ref{att}).
-This example would also work correctly if the assignments to \verb|y| were replaced with \verb|return| statements were placed directly in the branches of the conditional.
+We show two choices for these types on Lines 10 and 11.
+However, type inference correctly unifies these two types according to OpenCL's C99-derived rules governing numeric types (which are defined by the user in the \verb|OpenCL| module, as we will describe in Section \ref{att}).
+We test this programmatically on Lines 12 and 13.
+Note that this example would also work correctly if the assignments to \verb|y| were replaced with \verb|return| statements (in other words, the return value of a function is treated as an assignable for the purpose of type inference).
 
@@
 
-\subsection{Attribute and Extension Inference}
-In addition to type annotations, OpenCL introduces additional syntactic burdens.
-Developers must annotate functions that meet the requirements to be called from a host with the \verb|__kernel| attribute.
+\subsection{Annotation and Extension Inference}
+In addition to type annotations, OpenCL normally asks for additional annotations in a number of other situations.
+Users can annotate functions that meet certain requirements to be callable from the host with the \verb|__kernel| attribute.
 The \verb|Ace.OpenCL.OpenCL| backend is able to check these requirements and add this annotation automatically.
-Additionally, several types (notably, \verb|double|) and specialized functions require that an OpenCL extension be enabled with a \verb|#pragma|.
-The OpenCL backend can also  automatically infer many of these cases as well and add the appropriate \verb|#pragma| declaration, consistent with the principle of conciseness followed for the Ace design as a whole.
-An example of this can be seen in Listing \ref{mapout}.
+Several types (notably, \verb|double|) and specialized functions require that an OpenCL extension be enabled with a \verb|#pragma| when used.
+The OpenCL backend automatically detects many of these cases as well and adds the appropriate \verb|#pragma| declaration.
+An example of this feature can be seen on Line 5.1, where the use of the \verb|double| type triggers the insertion of an appropriate \verb|#pragma| automatically.
+Ace is designed to allow backends to observe the results of the type checking process to support this form of inference.
 
@@
 from ace.OpenCL import OpenCL
-import ast, ace.astx as astx
-
-plus = OpenCL.fn.from_string("""
+import ace.astx as astx
+
+plus = OpenCL.fn.from_str("""
 def plus(a, b):
@@
 \end{lstlisting}
-\caption{Metaprogramming with Ace, showing how to construct generic functions from strings and abstract syntax trees, and how to manipulate syntax trees at compile time.}
+\caption{[\texttt{listing7.py}] Metaprogramming with Ace, showing how to construct generic functions from both strings and abstract syntax trees, and how to manipulate syntax trees at compile-time.}
 \label{metaprogramming}
@@
 Metaprogramming refers to the practice of writing programs that manipulate other programs.
-There are a number of use cases for this technique, including domain-specific optimizations and code generation for programs with a repetitive structure that cannot easily be captured using available abstractions.
-OpenCL itself in particular relies on code generation as a fundamental mechanism, partially justifying its lack of support for higher-order programming \cite{pyopencl}.
-Ace supports programmatic compilation and higher-order constructs, as described above, and language extension, which we describe below, so several use cases for this feature have been eliminated.
-However, we demonstrate cases where this form of metaprogramming is useful in the case study described in Section \ref{casestudy}.
-
-An Ace function can be constructed from a string containing its source using the \verb|from_source| variant of \verb|fn|, as demonstrated in Listing \ref{metaprogramming} on Lines 4-7.
-It can also be constructed directly from a Python abstract syntax tree (AST), available via the Python standard \verb|ast| package, using the \verb|from_ast| variant of \verb|fn|, demonstrated on Line 10.
-The AST here is generated programmatically by transforming the syntax tree of the \verb|plus| function such that the argument \verb|b| becomes a constant, \verb|5|.
-This transformation, called \verb|specialize|, as well as several others are distributed with Ace in the \verb|ace.astx| module for convenience.
+There are a number of use cases for this technique, including domain-specific optimizations and code generation for programs with a repetitive structure that cannot easily be captured using available abstractions \cite{pyopencl}.
+OpenCL itself in particular relies on code generation as a fundamental mechanism, partially justifying its lack of support for higher-order programming.
+Ace supports programmatic compilation and higher-order constructs, as described above, and a flexible language extension mechanism, which we describe below, so several use cases for metaprogramming have been eliminated.
+However, cases where this form of metaprogramming include programmatic function specialization (Listing \ref{metaprogramming}) and modular simulation orchestration, described in Section \ref{clegans}.
+
+On Lines 4-7 of Listing \ref{metaprogramming}, an Ace function is constructed from a string containing its source using the \verb|from_source| variant of the \verb|OpenCL.fn| method.
+It can also be constructed directly from an abstract syntax tree (AST), as implemented by the Python standard \verb|ast| package, using the \verb|from_ast| variant of \verb|fn|, demonstrated on Line 10.
+The AST here is generated programmatically by copying the syntax tree of the \verb|plus| function with the argument, \verb|b|, eliminated and its uses replaced with a constant, \verb|5|.
+This transformation, called \verb|specialize|, as well as several others are distributed in the \verb|ace.astx| module for convenience.
 \subsection{Direct Invocation from Python}\label{direct}
@@
 from listing3 import map
-from listing4 import add5 # or Listing 7
+from listing4 import add5 # or from listing7
 
@@
 \end{lstlisting}
-\caption{A full OpenCL program using the \texttt{Ace.OpenCL} Python bindings, including data transfer and  invocation of a generic function, \texttt{map}, as a kernel without compiling it first.}
+\caption{[\texttt{listing8.py}] A full OpenCL program using the \texttt{Ace.OpenCL} Python bindings, including data transfer to and from a device and direct invocation of a generic function, \texttt{map}, as a kernel without explicit compilation.}
 \label{py}
@@
 As discussed in the Introduction, a common workflow for professional end-users involves the use of a high-level scripting language for overall workflow orchestration and small-scale data analysis and visualization, paired with a low-level language for performance-critical sections.
-Python is already widely used by professional end-users as the high-level scripting language.
-It has support for executing code written in low-level languages in several ways.
-Developers can call into native libraries using its foreign function interface (FFI) or a library like \verb|pycuda| for code compiled with CUDA, a proprietary language similar to OpenCL targeting nVidia GPUs \cite{pycuda}.
-
-The OpenCL language is designed specifically for this workflow, in that it exposes the compiler directly an API, rather than as a separate executable.
-Developers generate OpenCL source code as strings, compile it programmatically, then execute it using the run-time APIs that OpenCL also provides.
-Python bindings to the OpenCL APIs have been developed in a module called \verb|pyopencl| \cite{pyopencl}.
-
-Ace supports a similar workflow as an alternative to using \verb|acecc| as described above.
+Python is already widely used by professional end-users in the role of the high-level scripting language.
+It also has mature support for executing code written in low-level languages, in several ways.
+Developers can call into native libraries using its foreign function interface (FFI),  or by using a wrapper library like \verb|pycuda| for code compiled with CUDA, a proprietary language similar to OpenCL specifically targeting nVidia GPU hardware.
+Although CUDA's compilers are separate executables, the OpenCL language is tailored for this workflow, in that it exposes the compiler directly an API.
+The \verb|pyopencl| module exposes this API as well as the OpenCL memory management API to Python.
+With both \verb|pyopencl| and \verb|pycuda|, developers generate source code as strings, compile it programmatically, then execute it using the run-time APIs that each library provides \cite{pyopencl}.
+
+Ace supports this workflow as an alternative to using the \verb|acecc| compiler as described above to generate  source code directly.
+For the OpenCL backend, these bindings are exposed as a wrapper on top of \verb|pyopencl| called \verb|Ace.OpenCL.bindings|.
 Both generic functions and concrete functions written for a backend that supports the direct execution interface (thus far, \verb|OpenCL| and \verb|C99|) can be called like regular Python functions.
@@
 
-By way of comparison, the same program written using the OpenCL C API directly is an orders of magnitude larger and correspondingly more complex.
+By way of comparison, the same program written using the OpenCL C API directly is an order of magnitude larger and correspondingly more complex.
 A full implementation of the logic of \verb|map| written using the \verb|pyopencl| bindings and metaprogramming techniques as described in \cite{pyopencl} is twice as large and significantly more difficult to comprehend than the code we have shown thus far.
-Not shown are several additional conveniences, such as delegated kernel sizing and \verb|In| and \verb|Out| constructs that can reduce the size and improve the clarity of this code further -- the reader is referred to the language documentation for additional details.
+Not shown are several additional conveniences, such as delegated kernel sizing and \verb|In| and \verb|Out| constructs that can reduce the size and improve the clarity of this code further -- due to a lack of space, the reader is referred to the language documentation for additional details on these features.
 

### 2012-10-15 ace/ace.tex
ace: Finished revising Section II

@@
 There are a number of use cases for this technique, including domain-specific optimizations and code generation for programs with a repetitive structure that cannot easily be captured using available abstractions \cite{pyopencl}.
-OpenCL itself in particular relies on code generation as a fundamental mechanism, partially justifying its lack of support for higher-order programming.
+OpenCL in particular relies on code generation as a fundamental mechanism, which is cited as justification for its lack of support for higher-order programming.
 Ace supports programmatic compilation and higher-order constructs, as described above, and a flexible language extension mechanism, which we describe below, so several use cases for metaprogramming have been eliminated.
@@
 It can also be constructed directly from an abstract syntax tree (AST), as implemented by the Python standard \verb|ast| package, using the \verb|from_ast| variant of \verb|fn|, demonstrated on Line 10.
-The AST here is generated programmatically by copying the syntax tree of the \verb|plus| function with the argument, \verb|b|, eliminated and its uses replaced with a constant, \verb|5|.
-This transformation, called \verb|specialize|, as well as several others are distributed in the \verb|ace.astx| module for convenience.
+The AST here is generated programmatically by calling the \verb|specialize| function, which produces a copy of the syntax tree of the \verb|plus| function with the argument, \verb|b|, eliminated and its uses replaced with a constant, \verb|5|.
+This transformation as well as some others are distributed in \verb|ace.astx| for convenience.
 \subsection{Direct Invocation from Python}\label{direct}
@@
 from listing4 import add5 # or from listing7
-
-ctx = cl.Context.for_device(0, 0)
-d_in = ctx.to_device(np.ones((1024,)))
-d_out = ctx.alloc(like=d_in)
+cl.ctx = cl.Context.for_device(0, 0)
+
+input = np.ones((1024,))
+d_in = cl.to_device(input)
+d_out = cl.alloc(like=d_in)
 map(d_in, d_out, add5, 
     global_size=d_in.shape, local_size=(128,))
-out = ctx.from_device(d_out)
+out = cl.from_device(d_out)
+assert (out == a + 5).all()
 \end{lstlisting}
@@
 As discussed in the Introduction, a common workflow for professional end-users involves the use of a high-level scripting language for overall workflow orchestration and small-scale data analysis and visualization, paired with a low-level language for performance-critical sections.
-Python is already widely used by professional end-users in the role of the high-level scripting language.
-It also has mature support for executing code written in low-level languages, in several ways.
+Python is already widely used by professional end-users as a high-level scripting language and also features mature support for calling into code written in low-level languages.
 Developers can call into native libraries using its foreign function interface (FFI),  or by using a wrapper library like \verb|pycuda| for code compiled with CUDA, a proprietary language similar to OpenCL  specifically targeting nVidia GPU hardware.
-Although CUDA's compilers are separate executables, the OpenCL language is tailored for this workflow, in that it exposes the compiler directly an API.
+Although CUDA's compilers are separate executables on the system, the OpenCL language was designed for this workflow, in that it exposes the compiler directly as an API.
 The \verb|pyopencl| module exposes this API as well as the OpenCL memory management API to Python.
@@
 
-Ace supports this workflow as an alternative to using the \verb|acecc| compiler as described above to generate  source code directly.
+Ace supports this workflow as an alternative to using the \verb|acecc| compiler as described above to generate  source code directly from the shell.
 For the OpenCL backend, these bindings are exposed as a wrapper on top of \verb|pyopencl| called \verb|Ace.OpenCL.bindings|.
@@
 The first two arguments to \verb|map| are OpenCL buffers, generated using a simplified wrapper to the \verb|pyopencl| APIs on Lines 7-8.
-This wrapper associated type information with each buffer, and this information is used to implicitly compile \verb|map| as appropriate.
-That is, explicit calls to the \verb|compile| method, as we have been showing, are unnecessary when using this method of invocation.
-The final two keyword arguments on Line 10 are parameters that OpenCL requires for execution that determine the number of threads (called the {\em global size}) and the grouping of these threads (the {\em local size}).
+This wrapper associates type information with each buffer, similarly to \verb|numpy|, and this is used to implicitly compile \verb|map| as appropriate the first time it is called for any given combination of input types.
+Explicit calls to the \verb|compile| method, as we have been showing thusfar, are unnecessary if using this method of invocation.
+The final two keyword arguments on Line 10 are parameters that OpenCL requires for execution that determine the number of threads (called the {\em global size}) and thread grouping (the {\em local size}).
 
@@
 A full implementation of the logic of \verb|map| written using the \verb|pyopencl| bindings and metaprogramming techniques as described in \cite{pyopencl} is twice as large and significantly more difficult to comprehend than the code we have shown thus far.
-Not shown are several additional conveniences, such as delegated kernel sizing and \verb|In| and \verb|Out| constructs that can reduce the size and improve the clarity of this code further -- due to a lack of space, the reader is referred to the language documentation for additional details on these features.
+Not shown are several additional conveniences, such as delegated kernel sizing and \verb|In| and \verb|Out| constructs that can reduce the size and improve the clarity of this code further; due to a lack of space, the reader is referred to the language documentation for additional details on these features.
 

### 2012-11-21 ace/ace.tex
ace: Moved code samples into separate files.

@@
 
-Many of the barriers to adoption of new abstractions stem from the fact that new abstractions are often developed and distributed together with a new programming language.
+Many of the barriers to adoption faced by new abstractions stem from the fact that new abstractions are often developed and distributed together with a new programming language.
 Some prominent examples where this {\em language-oriented approach} has been followed include:
 \begin{itemize}
-\item UPC \cite{upc}, CAF \cite{caf}, X10 \cite{x10} and Chapel \cite{chapel}, built around first-class support for a partitioned global address space (PGAS) programming model; 
-\item Erlang \cite{erlang}, Charm++ \cite{charm}, Occam \cite{occam} and Axum \cite{axum}, built around first-class support for message passing;
-\item Cilk \cite{cilk} and NESL \cite{nesl}, built around first-class support for nested data parallelism; and
-\item Add something about automatic parallelization of functional programs.
+\item UPC, CAF, X10 and Chapel, built around first-class support for a partitioned global address space (PGAS) programming model; 
+\item Erlang, Charm++, Occam and Axum, built around first-class support for message passing;
+\item Cilk and NESL, built around first-class support for nested data parallelism; and
+\item {\color{red} Add something about automatic parallelization of functional programs.}
+\item {\color{red} Add something about CSP like Go.}
 \end{itemize}
 
-There is evidence suggesting that developers would prefer first-class support for a parallel abstraction over a library-based solution, if all else is held constant \cite{ppfirstclass}.
-Unfortunately, adopting a specialized research language comes with a host of ancillary challenges.
-Learning a new language can be time-consuming, new languages often lack well-developed libraries and tools, and it can be difficult to port large legacy codebases.
-Indeed, the most widely adopted parallel programming abstractions have been those that are distributed as libraries for an existing, widely-used language, such as MapReduce, Global Arrays and MPI.
+There is indeed evidence suggesting that developers would prefer first-class support for a parallel abstraction over a library-based solution, when all else is held constant \cite{ppfirstclass}.
+Unfortunately, adopting a specialized research language comes with a host of ancillary challenges \cite{adoptionTR,socioplt}.
+Indeed, the most widely {\em adopted} parallel programming abstractions have been those that are distributed as libraries for an existing, widely-used language, such as MapReduce, Global Arrays and MPI.
 This {\em library-oriented approach} has proven more practical because it allows developers to progressively introduce parallelism into existing programs, leverage a familiar and well-developed general-purpose programming language when not directly invoking parallel primitives, and utilize different abstractions within a single program without encountering the friction inherent to language boundaries.
 
-Ace aims to solve this problem by allowing developers to define and distribute new {\em first-class language constructs}, in a safely composable manner, directly within user libraries, via a mechanism we introduce called {\em active typechecking and translation (\ATT)}.
-In particular, \ATT~is a mechanism that allows developers to  extend the semantics of a language that has a fixed but flexible grammar.
-The Ace grammar is a large subset of Python's  and we show that it is flexible enough that we are able to define all of the primitives of the C99 and OpenCL languages in a natural manner, aided by a form of type inference.
-These primitives can then be supplemented with higher-level user-defined primitives, examples of which we discuss in Section \ref{abstractions}.
-Notably, all of these primitives are imported via the normal library loading mechanism; they are not defined within domain-specific languages, nor as extensions for a particular compiler for a language.
+Ace aims to bridge these two approaches by allowing developers to define and distribute new {\em first-class language constructs} directly within user libraries, using a mechanism we introduce called {\em active typechecking and translation (\ATT)}.
+Specifically, \ATT~allows developers to  extend the semantics of a language that has a fixed, but flexible, grammar.
+The Ace grammar is a nearly-complete subset of Python's  and we show that it is flexible enough to support all of the primitives of the C99 and OpenCL languages in a natural manner, when aided by a form of type inference that we describe.
+We further supplement these primitives with higher-level user-defined primitives, examples of which we discuss in Section \ref{abstractions}.
+Notably, all of these primitives are imported via the normal library loading mechanism; they are not defined within domain-specific languages, nor as extensions for a particular compiler implementation.
 
@@
 Indeed, the language itself is implemented entirely as a Python library, and many existing tools, such as editor modes, documentation tools and style checkers, can be used without modification.
-Ace functions are generated programmatically and can be called directly from Python, with numeric arrays as arguments, so developers can integrate Ace immediately within their existing workflows to define portions of code that may benefit from static compilation and support for parallelism.
+Ace functions are generated programmatically and can be called directly from Python, with numeric arrays as arguments, so developers can integrate Ace immediately within their existing workflows for portions of code that may benefit from static compilation and support for parallelism.
 We demonstrate the practicality of Ace in Section \ref{clegans} by describing an Ace-based framework that supports specifying and orchestrating families of large-scale neurobiological circuit simulations on clusters of GPUs.
@@
 \begin{codelisting}
-\begin{lstlisting}
-from ace.OpenCL import OpenCL, printf
-
-print "Hello, compile-time world!"
-
-@OpenCL.fn
-def main():
-    printf("Hello, run-time world!")
-main = main.compile()
-
-print "Goodbye, compile-time world!"
-\end{lstlisting}
-\caption{\texttt{hello.py}}
+\lstinputlisting{hello.py}
+\caption{\texttt{[hello.py]} A basic Ace program demonstrating the two-phase structure of Ace programs and libraries.}
 \label{hello}
@@
 \begin{lstlisting}[style=Bash]
-$ `acecc hello.py`
+$ `acec hello.py`
 Hello, compile-time world.
@@
 }\end{lstlisting}
-\caption{Compiling \texttt{hello.py} using \texttt{acecc}.}
+\caption{Compiling \texttt{hello.py} using the \texttt{acec} compiler.}
 \label{helloout}
@@
 
-\section{Ace for Professional End-Users}
+\section{Structure and Usage}
 %Ace is an {\em actively-typed, language-integrated compilation environment}. Let us begin by defining each of these terms. 
@@
 \subsection{Example 1: Hello, World!}
-To demonstrate the structure of programs and libraries in Ace, let us begin in Listing \ref{hello} with a variant of the canonical ``Hello, World!'' example written using the OpenCL module distributed with Ace.
+To demonstrate the somewhat unconventional structure of Ace programs and libraries, we begin in Listing \ref{hello} with a variant of the canonical ``Hello, World!'' example written using the \verb|OpenCL| module.
 We note that this module, which we use throughout the paper in examples, is simply a library like any other.
 The core of Ace gives no special treatment to it; it is distributed with Ace for convenience.
+Aspects of its implementation will be described in Section \ref{att}.
 
@@
 The distinction between generic and concrete functions will be explained in our next example.
-For now, we note that only concrete functions are exported when running the Ace compiler, called \verb|acecc|, shown operating at the shell in Listing \ref{helloout}.
-The \verb|acecc| compiler operates as follows:
+For now, we note that only concrete functions are exported when running the Ace compiler, called \verb|acec|, shown operating at the shell in Listing \ref{helloout}.
+The \verb|acec| compiler operates as follows:
 \begin{enumerate}
@@
 
-Using the \verb|acecc| executable is simply a convenience; the generated code is also available as an attribute called of the concrete function immediately after it has been defined, accessed as \verb|main.code|.
+Using the \verb|acec| executable is simply a convenience; the generated code is also available as an attribute of the concrete function immediately after it has been defined, accessed as \verb|main.code|.
 We will show in Section \ref{pyopencl} that, for backend languages with Python bindings, such as OpenCL and C, generic functions can be executed directly, without the intermediate compilation step, if desired.
@@
 \begin{codelisting}
-\begin{lstlisting}
-from ace.OpenCL import OpenCL, get_global_id
-
-@OpenCL.fn
-def map(input, output, f):
-    gid = get_global_id(0)
-    output[gid] = f(input[gid])
-\end{lstlisting}
-\caption{[\texttt{listing3.py}] A generic data-parallel higher-order map function written using OpenCL user module.}
+\lstinputlisting{listing3.py}
+\caption{[\texttt{listing\ref{map}.py}] A generic data-parallel higher-order map function written using the OpenCL user module.}
 \label{map}
@@
 \begin{codelisting}
-\begin{lstlisting}
-from listing3 import map
-from ace.OpenCL import global_ptr, double, int
-
-@OpenCL.fn
-def add5(x):
-    return x + 5
-
-A = global_ptr(double); B = global_ptr(int)
-map_add5_dbl = map.compile(A, A, add5.ace_type)
-map_add5_int = map.compile(B, B, add5.ace_type)
-\end{lstlisting}
-\caption{[\texttt{listing4.py}] The generic \texttt{map} function compiled to map the \texttt{add5} function over two  types of input.}
+\lstinputlisting{listing4.py}
+\caption{[\texttt{listing\ref{mapadd5dbl}.py}] The generic \texttt{map} function compiled to map the \texttt{add5} function over two  types of input.}
 \label{mapadd5dbl}
@@
 \begin{codelisting}
-\begin{lstlisting}[style=OpenCL]
-#pragma OPENCL EXTENSION cl_khr_fp64 : enable
-
-double add5(double x) {
-    return x + 5;
-}
-
-__kernel void map_add5_dbl(
-  __global double* input, 
-  __global double* output) 
-{
-    size_t gid;
-    gid = get_global_id(0);
-    output[gid] = add5(input[gid]);
-}
-
-int add5__1(int x) {
-    return x + 5;
-}
-
-__kernel void map_add5_int(
-  __global int* input,
-  __global int* output)
-{
-    size_t = gid;
-    gid = get_global_id(0);
-    output[gid] = add5__1(input[gid]);
-}
-\end{lstlisting}
-\caption{[\texttt{listing4.cl}] The OpenCL code generated by running \texttt{acecc listing4.py}.}
+\lstinputlisting[style=OpenCL]{listing5.cl}
+\caption{[\texttt{listing\ref{mapout}.cl}] The OpenCL code generated by running \texttt{acec listing2.py}.}
 \label{mapout}
@@
 
-Running \verb|acecc listing3.py| would produce no output, however.
+Running \verb|acec listing3.py| would produce no output, however.
 To create a {\em concrete function} (an instance of the class \verb|ace.ConcreteFn|) that can be emitted by the compiler, types must be assigned to each of the arguments of any externally callable functions.
@@
 \begin{codelisting}
-\begin{lstlisting}
-from ace.OpenCL import OpenCL, int, double, long
-
-@OpenCL.fn
-def threshold_scale(x, scale):
-    if x <= 0:
-        y = 0
-    else:
-        y = scale * x
-    return y
-f = f.compile(int, double)
-g = g.compile(int, long)
-assert f.return_type == double
-assert g.return_type == long
-\end{lstlisting}
-\caption{A function demonstrating whole-function type inference when multiple values with differing types are assigned to a single variable.}
+\lstinputlisting{listing6.py}
+\caption{\texttt{[listing6.py]} A function demonstrating whole-function type inference when multiple values with differing types are assigned to a single variable.}
 \label{inference}
@@
 \begin{codelisting}
-\begin{lstlisting}[commentstyle=\color{mauve}]
-from ace.OpenCL import OpenCL
-import ace.astx as astx
-
-plus = OpenCL.fn.from_str("""
-def plus(a, b):
-    return a + b
-""")
-
-add5_ast = astx.specialize(plus.ast, b=5)
-add5 = OpenCL.fn.from_ast(add5_ast)
-\end{lstlisting}
+\lstinputlisting[commentstyle=\color{mauve}]{listing7.py}
 \caption{[\texttt{listing7.py}] Metaprogramming with Ace, showing how to construct generic functions from both strings and abstract syntax trees, and how to manipulate syntax trees at compile-time.}
@@
 \begin{codelisting}
-\begin{lstlisting}
-import numpy as np
-import ace.OpenCL.bindings as cl
-from listing3 import map
-from listing4 import add5 # or from listing7
-cl.ctx = cl.Context.for_device(0, 0)
-
-input = np.ones((1024,))
-d_in = cl.to_device(input)
-d_out = cl.alloc(like=d_in)
-map(d_in, d_out, add5, 
-    global_size=d_in.shape, local_size=(128,))
-out = cl.from_device(d_out)
-assert (out == a + 5).all()
-\end{lstlisting}
-\caption{[\texttt{listing8.py}] A full OpenCL program using the \texttt{Ace.OpenCL} Python bindings, including data transfer to and from a device and direct invocation of a generic function, \texttt{map}, as a kernel without explicit compilation.}
+\lstinputlisting{listing8.py}
+\caption{[\texttt{listing\ref{py}.py}] A full OpenCL program using the \texttt{Ace.OpenCL} Python bindings, including data transfer to and from a device and direct invocation of a generic function, \texttt{map}, as a kernel without explicit compilation.}
 \label{py}
@@
 
-Ace supports this workflow as an alternative to using the \verb|acecc| compiler as described above to generate  source code directly from the shell.
+Ace supports this workflow as an alternative to using the \verb|acec| compiler as described above to generate  source code directly from the shell.
 For the OpenCL backend, these bindings are exposed as a wrapper on top of \verb|pyopencl| called \verb|Ace.OpenCL.bindings|.
@@
 However, Ace gives no preferential treatment to this module; it is implemented entirely using the user-facing mechanisms described in this section.
-
-Most programming languages are {\em monolithic} -- the set of available primitives is determined by the language designers, and users must combine these primitives to produce desired behaviors.
-Although a number of very flexible primitives have been developed (e.g.
-object systems), these have proven insufficient in specialized domains where developers and researchers need fine control over how certain operations are type checked and translated.
+A C99 module has also been developed (which, due to its similarity to OpenCL, shares many of its implementation details), but we do not discuss it further in this paper.
+Extensions complementing these are described in Section C below.
+
+Most programming languages are {\em monolithic} -- the set of available primitives is determined by the language designers, and users must combine these primitives to produce any desired run-time behavior.
+Although many very general primitives have been developed (e.g.
+object systems and algebraic datatypes), these can be insufficient in specialized domains where developers and researchers need fine control over how certain operations are type checked and translated.
 High-performance computing is an example of such a field, since both correctness and performance have been difficult to achieve in general, and are topics of active research.
-As discussed in the Introduction, the proliferation of parallel programming languages, rather than library-based abstractions, indicates that researchers often find the abstractions in existing languages insufficient for their needs.
-
-In view of these issues, Ace is fundamentally {\em extensible}, rather than monolithic.
-Users can introduce new primitive types and operations, fully controlling how they are type checked and translated.
-The target of translation can also be specified modularly by the user.
-Because Ace libraries can contain compile-time logic, written in the metalanguage as described in the previous section, these primitive definitions can be distributed within normal modules.
+As discussed in the Introduction, the proliferation of parallel programming languages, rather than library-based abstractions, indicates that researchers often find the abstractions available in existing languages insufficient for their needs.
+
+To address these use cases, Ace has been designed to be fundamentally {\em extensible}, rather than monolithic.
+Users can introduce new primitive types and operations and fully control how they are typechecked and translated.
+The backend target of translation can also be specified modularly by the user.
+Because Ace libraries can contain compile-time logic, written in the metalanguage as described in the previous section, these primitive definitions can be distributed as modules, rather than as extensions to particular compilers or domain-specific languages.
 
 \subsection{Active Typechecking and Translation (\ATT)}
-
-When the compiler encounters an expression, it must first verify that it typechecks, then translate it to produce code in a target language.
-Rather than containing fixed logic for this, however, the Ace compiler defers this responsibility to the {\it type} of a subexpression whenever possible, according to a fixed {\em dispatch protocol} for each syntactic form.
+We now explain how new primitive types and operators can be defined in Ace.
+When the compiler encounters an expression, such as \verb|input[gid]|, it must first verify its validity by assigning it a type, then translate the expression to produce an expression in a target language.
+Rather than containing fixed logic for this, however, the Ace compiler defers this responsibility to the {\it type} of a subexpression, such as \verb|input|, whenever possible, according to a fixed {\em dispatch protocol} for each syntactic form.
 Below are examples of the rules that comprise the Ace dispatch protocol.
-Due to space constraints, we do not list the entire dispatch protocol, which contains a rule for each possible syntactic form.
+Due to space constraints, we do not list the entire dispatch protocol, which contains a rule for each possible syntactic form in the language.
 \begin{itemize}
@@
 \item Responsibility over {\bf attribute access} (\texttt{obj.attr}) and {\bf subscript access}, (\texttt{obj[idx]}) is handed to the type assigned to \texttt{obj}.
+\item {\color{red} Talk about multiple assignment here I think}
 \end{itemize}
+\begin{codelisting}
+\lstinputlisting{listing9.py}
+\caption{\texttt{[listing9.py]} A portion of the implementation of OpenCL pointer types implementing subscripting logic using the Ace extension mechanism.}
+\label{pointers}
+\end{codelisting}
 
 \subsubsection{Active Typechecking}
-During the typechecking phase, the type of the primary operand, as determined by the dispatch protocol, is responsible for assigning a type to the expression as a whole.
+During the typechecking phase, the type of the primary operand, as determined by this dispatch protocol, is responsible for assigning a type to the expression as a whole.
 Let us consider the \verb|map| function from Listing \ref{map} once again.
@@
 In this case, the expression is of the \verb|Subscript| form, so the compiler calls \verb|A.resolve_Subscript|.
-
-\begin{codelisting}
-\begin{lstlisting}
-import ace, ace.astx as astx
-
-class PtrType(ace.Type):
-  def __init__(self, T, addr_space):
-    self.target_type = T
-    self.addr_space = addr_space
-
-  def resolve_Subscript(self, context, node):
-    slice_type = context.resolve(node.slice)
-    if isinstance(slice_type, IntegerType):
-      return self.target_type
-    else:
-      raise TypeError('<error message>', node)
-
-  def translate_Subscript(self, context, node):
-    value = context.translate(node.value)
-    slice = context.translate(node.slice)
-    return astx.copy_node(node, 
-      value=value, slice=slice,    
-      code=value.code + '[' + slice.code + ']')
-
-  # ...
-
-class GlobalPtrType(PtrType):
-  def __init__(self, T):
-    PtrType.__init__(self, T, '__global')
-\end{lstlisting}
-\caption{A portion of the implementation of OpenCL pointer types implementing subscripting logic using the Ace extension mechanism.}
-\label{pointers}
-\end{codelisting}
 
@@
 \begin{codelisting}
-\begin{lstlisting}
-# Create the root node of the simulation
-sim = Simulation(ctx, n_timesteps=10000)
-neurons = ReducedLIF(sim, count=N, tau=20.0)
-e_synapse = ExponentialSynapse(neurons, 'ge',
-    tau=5.0, reversal=60.0)
-probe = StateVariableProbeCopyback(e_synapse)
-\end{lstlisting}
-\caption{An example of a nested simulation tree, showing that specifying a simulation is both simple and modular.}
+\lstinputlisting{listing10.py}
+\caption{\texttt{[listing10.py]} An example of a nested simulation tree, showing that specifying a simulation is both simple and modular.
+The first argument to the constructor specifies each node's parent.}
 \label{spec}
@@
 \begin{codelisting}
-\begin{lstlisting}
-class SpikingModel(Model):
-  """Base class for spiking neuron models."""
-  def in_model_code(self, g):
-    "idx_state = idx_model + count*" << g
-    "(realization_n - realization_start)" << g
-    self.insert_hook("read_incoming", g)
-    self.insert_hook("read_state", g)
-    self.insert_hook("calculate_inputs", g)
-    self.insert_hook("state_calculations", g)
-    self.insert_hook("spike_processing", g)
-  # ...
-\end{lstlisting}
-\caption{An example of a hook that inserts code and also inserts new, nested hooks for downstream simulation nodes  below that.}
+\lstinputlisting{listing11.py}
+\caption{\texttt{[listing11.py]} An example of a hook that inserts code and also inserts new, nested hooks for downstream simulation nodes  below that.}
 \label{impl}

### 2012-11-21 ace/hello.py

@@ -1,0 +1,10 @@
+from ace.OpenCL import OpenCL, printf
+ 
+print "Hello, compile-time world!"
+ 
+@OpenCL.fn
+def main():
+    printf("Hello, run-time world!")
+main = main.compile()
+
+print "Goodbye, compile-time world!"

### 2012-11-21 ace/listing11.py

@@ -1,0 +1,11 @@
+class SpikingModel(Model):
+  """Base class for spiking neuron models."""
+  def in_model_code(self, g):
+    "idx_state = idx_model + count*" << g
+    "(realization_n - realization_start)" << g
+    self.insert_hook("read_incoming", g)
+    self.insert_hook("read_state", g)
+    self.insert_hook("calculate_inputs", g)
+    self.insert_hook("state_calculations", g)
+    self.insert_hook("spike_processing", g)
+  # ...

### 2012-11-21 ace/listing4.py

@@ -1,0 +1,10 @@
+from listing3 import map
+from ace.OpenCL import global_ptr, double, int
+
+@OpenCL.fn
+def add5(x):
+    return x + 5
+
+A = global_ptr(double); B = global_ptr(int)
+map_add5_dbl = map.compile(A, A, add5.ace_type)
+map_add5_int = map.compile(B, B, add5.ace_type)

### 2012-11-21 ace/listing8.py

@@ -1,0 +1,16 @@
+import numpy as np
+import ace.OpenCL.bindings as cl
+from listing1 import map
+from listing2 import add5
+
+cl.ctx = cl.Context.for_device(0, 0)
+
+input = np.ones((1024,))
+d_in = cl.to_device(input)
+d_out = cl.alloc(like=d_in)
+
+map(d_in, d_out, add5, 
+    global_size=d_in.shape, local_size=(128,))
+
+out = cl.from_device(d_out)
+assert (out == a + 5).all()

### 2012-11-21 ace/listing9.py

@@ -1,0 +1,26 @@
+import ace, ace.astx as astx
+
+class PtrType(ace.Type):
+  def __init__(self, T, addr_space):
+    self.target_type = T
+    self.addr_space = addr_space
+        
+  def resolve_Subscript(self, context, node):
+    slice_type = context.resolve(node.slice)
+    if isinstance(slice_type, IntegerType):
+      return self.target_type
+    else:
+      raise TypeError('<error message>', node)
+       
+  def translate_Subscript(self, context, node):
+    value = context.translate(node.value)
+    slice = context.translate(node.slice)
+    return astx.copy_node(node, 
+      value=value, slice=slice,    
+      code=value.code + '[' + slice.code + ']')
+      
+  # ...
+   
+class GlobalPtrType(PtrType):
+  def __init__(self, T):
+    PtrType.__init__(self, T, '__global')

### 2013-01-09 thesis-proposal/background.tex
Initial thesis proposal taken from NSF grant

\section{Background}
We review existing approaches that are available to researchers and domain experts who wish to develop novel constructs below.

\subsection{Language-Oriented Approaches}
\subsubsection{Language Frameworks}
A number of tools have been developed to assist with this task of developing new languages and their associated tools, including compiler generators, language workbenches and domain-specific language frameworks (see \cite{fowler2010domain} for a review). In some cases, these tools allow language features to be defined modularly and composed differentially to produce a variety of different languages. To our knowledge, none of these mechanisms guarantee that any combination of features can be safely composed, nor do they guarantee safety properties about the resulting languages. User-defined code is still ultimately compiled against a particular language, and because language composition cannot be automatic or guaranteed safe, interoperability with code written in other languages is thus limited by the difficulties described above despite the modular construction.

\subsubsection{Extensible Compilers and Tools}
A related methodology is to implement language features as compiler and tool extensions directly. A number of extensible compilers have been developed to support this approach (see \cite{clements2008comparison}). As with language frameworks, this approach can lead to \emph{de facto} implementation of a new language, since user libraries must be compiled against a particular combination of extensions, and these extensions cannot be guaranteed to compose in general. Moreover, a language may have multiple competing compilers and other tools. By relying on implementation-specific features of a single tool to define core semantics and behaviors, the clean conceptual separation between languages and tools is broken, leading to compatibility issues and hard-to-anticipate behaviors for user code. We argue that this approach should be considered harmful.

\subsection{Library-Oriented Approaches}
\subsubsection{Embedded DSLs}
Embedded domain-specific languages are languages that creatively repurpose existing language constructs to create interfaces that resemble those of a distinct language. In languages with rich type systems, such as Haskell, this approach can be quite successful (e.g. \cite{svensson2011obsidian}). Ultimately, however, this approach is limited by the host system, and as discussed in Section 2, thus limits experts who want to express particularly novel constructs, relative to the host language.

\subsubsection{Term Rewriting Systems}
Many languages and tools allow developers to rewrite expressions according to custom rules. These can broadly be classified as {\it term rewriting systems}. Macro systems, such as those characteristic of the LISP family of languages \cite{mccarthy1978history}, are the most prominent example. Some compile-time metaprogramming systems also allow users to manipulate syntax trees (e.g. MetaML \cite{Sheard:1999:UMS}), and external rewrite systems also exist for many languages. These systems are expressive if used correctly, but verifying correctness and non-interference properties is difficult for the same reason. Manipulating source trees directly is a complex task, even in languages with simple grammars like LISP. Finally, term rewriting systems focus on rewriting terms to support alternative language semantics but do not intrinsically support extensions that cover the full programming system.

\subsubsection{Active Libraries}
Active libraries, as proposed by Czarnecki et al. \cite{activelibraries}, ``are not passive collections of routines or objects, as are traditional libraries, but take an active role in generating code''. They go on to suggest a number of areas in which the library could interact with the programming system, including optimizing code, checking source code for correctness, reporting domain-specific errors and warnings, and ``rendering domain-specific textual and non-textual program representations and for interacting with such representations''.

This paper was largely a proposal. The concrete implementations of this concept have largely been term rewriting systems described above. One prominent example within the active libraries literature is Blitz++, which uses the C++ template expansion system as a metalanguage to support optimizations of array operations. An example of tool support comes from the Tau package for tuning and analyzing parallel programs. Tau allows libraries to instrument themselves declaratively to hide internal details and complex internal representations from users. A more extensive system with support for active libraries is Xroma. Xroma allows users to provide annotations that intercept compilation of a component at various stages of compilation to support rewriting, custom error handling and custom error checking. The approach taken by Xroma, by still requiring direct syntax manipulation and allowing arbitrary interception, is flexible but not compositional, and generally more complex than may be necessary. There also does not appear to be a clear, well-founded theoretical foundation for this approach, nor for active libraries in general.

%\subsection{Type-Level Computation} %Haskell, Ur and $\Omega$mega
%System XX with simple case analysis provides the basis of type-level computation in Haskell (where type-level functions are called type families \cite{Chakravarty:2005:ATC}). Ur uses type-level records and names to support typesafe metaprogramming, with applications to web programming \cite{conf/pldi/Chlipala10}. $\Omega$mega adds algebraic data types at the type-level, using these to increase the expressive power of algebraic data types at the expression level \cite{conf/cefp/SheardL07}. Dependently-typed languages blur the traditional phase separation between types and expressions, so type-level computation is often implicitly used (though not always in its most general form, e.g. Deputy \cite{conf/icfp/ChenX05}, ATS \cite{conf/esop/ConditHAGN07}.)

%\subsubsection{Run-Time Indirection}
%{\it Operator overloading} \cite{vanWijngaarden:Mailloux:Peck:Koster:Sintzoff:Lindsey:Meertens:Fisker:acta:1975} and {\it metaobject dispatch} \cite{Kiczales91} are run-time protocols that translate operator invocations into function calls. The function is typically selected according to the type or value of one or more operands. These protocols share the notion of {\it inversion of control} with type-level specification. However, type-level specification is a {\it compile-time} protocol focused on enabling specialized verification and implementation strategies, rather than simply enabling run-time indirection.

%\subsection{Specification Languages}
%Several {\it specification languages} (or {\it logical frameworks}) based on these theoretical formulations exist, including the OBJ family of languages (e.g. CafeOBJ \cite{Diaconescu-Futatsugi01}). They provide support for verifying a program against a language specification, and can automatically execute these programs as well in some cases. The  language itself specifies which verification and execution strategies are used.
%
%Type-determined compilation takes a more concrete approach to the problem, focusing on combining {\it implementations} of different\- logics, rather than simply their specifications. In other words, it focuses on combining {\it type checkers} and {\it implementation strategies} rather than more abstract representations of a language's type system and dynamic semantics. In Section 4, we outlined a preliminary approach based on proof assistant available for the type-level language to unify these approaches, and we hope to continue this line of research in future work.
%

### 2013-01-09 thesis-proposal/education.tex
Initial thesis proposal taken from NSF grant

\section{Education}
Actively-typed programming systems are useful for researchers and domain experts because they allow them to develop and distribute new constructs modularly. This modularity is also useful to educators who wish to incrementally introduce advanced language features to students, as well as to integrate more helpful error messages and diagnostics that may be less useful to professional programmers and more advanced students.

Our proposed work in this area will center largely around using Ace in the classroom. As of this writing, Ace is best suited for teaching low-level programming because of the library-based implementations of C99 and OpenCL. At CMU, the introductory course on Principles of Imperative Programming teaches a language called C0, which is a significant simplification of the C language, augmented with some features to improve safety and allow users to explicit include pre-conditions and post-conditions on functions. This kind of development effort is precisely suited to an actively-typed language like Ace -- features can be implemented in libraries, which can be introduced incrementally, and new features can be added to complement existing libraries easily. Thus, we plan on evaluating the suitability of Ace for this course and may expand to similar efforts at other institutions.

Another course, Advanced Computer Graphics, teaches GPU programming to students using a combination of the CUDA and OpenCL languages. We have presented Ace as a guest lecture to this class in the past, and the feedback has been overwhelmingly positive. We are discussing incorporating Ace as a replacement for OpenCL, due largely to its simplicity from the end-user perspective, in future editions of this course. This opportunity will allow us to analyze the learning curve for Ace, and compare it to past editions of the course reasonably directly. We will also be able to instrument the compiler to gather detailed, anonymous data to complement the studies we discuss in Section 5.1.

### 2013-01-09 thesis-proposal/preliminary.tex
Initial thesis proposal taken from NSF grant

\section{Preliminary Work}
We have started implementing a form of active typechecking and translation, as outlined above, by developing an actively-typed programming language embedded within Python called Ace. In addition, we have demonstrated and empirically evaluated active code completion support for Java, implemented using class annotations and a  plugin for the Eclipse editor called Graphite.

\subsection{Ace: Active Typechecking and Translation}
Ace is an actively-typed programming language with a fixed syntax, shared with the Python language, but user-extensible semantics specified using an implementation of the active typing mechanism described above. The type-level language of Ace is Python, and types are type-level instances of Python classes implementing a provided interface, \verb|ace.Type|. The compiler selectively invokes the methods of these instances when type checking and translating expressions, a mechanism we call {\em active type-checking and translation (AT\&T)}. We have demonstrated the basic practicality of this mechanism by implementing primitives from several widely-known low-level languages, including much of C99 and the entirety of OpenCL, as active libraries. 
\begin{codelisting}
\lstinputlisting{listing4.py}
\caption{The generic \texttt{map} function compiled to map the \texttt{add5} function over two  types of input.}
\label{mapadd5dbl}
\end{codelisting}

\begin{codelisting}
\lstinputlisting{listing9.py}
\caption{A portion of the implementation of OpenCL pointer types implementing subscripting logic using the Ace extension mechanism.}
\label{pointers}
\end{codelisting}

An example showing usage of Ace with the OpenCL active library is shown in Listing \ref{mapadd5dbl}. Here, the top-level of the file is type-level code. On line 13, two instances of an active type, corresponding to the OpenCL types \verb|__global double*| and \verb|__global int*|, are created (to emphasize, at compile-time). The \emph{generic functions} defined on lines 4-11 are then programmatically specialized with these types assigned to the input arguments on lines 12-13, triggering active typechecking and translation mechanism when operations on these arguments are encountered (e.g. line 7).

Listing \ref{pointers} shows the portion of the implementation of the \verb|GlobalPtrType| type family (which \verb|A| and \verb|B| in Listing 1 are members of) that is responsible for the subscripting operation on line 7. The \verb|resolve_Subscript| method is responsible for checking that the subscript is an integer type, returning the appropriate type for the expression as a whole -- the target type of the pointer -- if so or raising an appropriate error if not. The \verb|translate_Subscript| function is then responsible for translating the subscript expression into the target language. Here, our target language is OpenCL (that is, we have simply lifted a target language construct into Ace directly), so the translation is straightforward. In general, however, this method may contain sophisticated code generation logic. These two methods correspond to the $f_{\text{resolve-X}}$ and $f_{\text{compile-X}}$ functions described above.

In addition to AT\&T, Ace supports more general forms of metaprogramming, and functions can be launched directly from Python with standard numeric data structures as arguments. Using Ace, we designed a scientific simulation framework that allows users to modularly specify, compile and orchestrate the execution of parameterized families of scientific simulations on clusters of GPUs. This framework has been used to successfully conduct large-scale, high-performance neuroscience simulations, providing initial evidence that Ace is useful not only as a foundational tool for researchers, but also for the practice of high-performance computing today.

\subsection{Graphite: Active Code Completion}\label{graphite}
Graphite demonstrates an edit-time aspect of active types, corresponding to the $f_{\text{editor}}$ function above, in the context of an existing language, Java, and an existing editor, Eclipse. A paper describing and evaluating this work which was recently published at ICSE 2012 \cite{omar2012active}. 
In that paper, we introduced {\em active code completion}, a mechanism that allows library developers to associate interactive and highly-specialized code generation interfaces, called {\em palettes}, directly with types using Java's annotation system and HTML5 for implementation. A simple example of such a palette and its operation is shown in Figure \ref{color}.

Using several empirical methods, we examined the contexts in which such a system could be useful, described the design constraints governing the system architecture as well as particular code completion interfaces, and detailed the design of our implementation, Graphite. Using Graphite, we implemented a more sophisticated palette for writing regular expressions and conducted a small pilot study that showed that this kind of domain-specific interface is useful for professional developers working in that domain.

\begin{figure*}\label{color}
\begin{center}
\includegraphics[width=40pc]{color_palette.png}\end{center}
\caption{(a) An example code completion palette associated with the \texttt{Color} class. (b) The source code generated by this palette.}
\end{figure*}

### 2013-01-09 thesis-proposal/problem.tex
Initial thesis proposal taken from NSF grant

\section{Problem}

Designing and implementing a programming language and its supporting tools (collectively, a \emph{programming system}) that simultaneously supports a natural programming style, helps users find and fix errors early, and performs well across diverse problem domains remains a grand challenge in computing. Indeed, it is now widely acknowledged that no simple set of constructs will be able to handle all of the challenges modern computing faces. Consistent with this view, a large number of researchers and domain experts in both research and industry are actively engaged in the design, implementation and analysis of a variety of novel language-based abstractions, compilation strategies and tools targeted at different problem domains.

%In implementing a new feature or abstraction, these experts can take one of several approaches. 
To realize a new feature or abstraction, these experts often take a \emph{language-oriented approach}, whereby the new constructs are developed and distributed together with a new programming language. The flexibility of this approach, coupled with support from tools like compiler generators, has led to a proliferation of specialized languages, each built around a particular set of privileged constructs. Unfortunately, a key issue limiting the practicality of this approach occurs at the boundaries between languages -- specialized constructs particular to one language cannot always be safely and naturally expressed in another, so building a program out of components written in different specialized languages can often prove difficult.

One common strategy that partially addresses this problem is to target an established language and runtime, such as the Java Virtual Machine (JVM), and support a superset of its features. This only solves the interoperability problem in one direction. While calling from the new language to the old language becomes straightforward, calls in the other direction, or between languages sharing a common target, remain restricted by the constructs available in the common language. This issue is illustrated in Figure \ref{approaches}a. This restriction can  constrain the design of new languages to only include constructs that can be expressed reasonably naturally in the target language. More novel innovations are often difficult to translate into familiar forms, however. For instance, the module system in F\# is severely constrained relative to the module system in its predecessor, OCaml, due to the need for interoperability with other CLI languages.

Stepping back, it is clear that a \emph{library-oriented approach}, shown in Figure \ref{approaches}b, where new constructs are implemented entirely within a single host language as libraries, should be preferred when possible.
% -- library-based constructs cannot interfere with one another (assuming a suitable namespacing standard), libraries benefit from common distribution infrastructure and tool support, and externally-accessible library interfaces need not be restricted to a fixed common subset of all available constructs (e.g. the object system of the JVM or CLI).  We diagram this key latter distinction in Figure \ref{approaches}. 
Of course, this is very often \emph{not} possible today because from the perspective of a library developer, the programming system is a largely inflexible, \emph{monolithic} entity. That is, the language's syntax, type checking rules and semantics are fixed in advance, the compiler is a black box implementation of these rules, and the tools operate according to domain-agnostic protocols that use only the declared structure of the program being developed as input. To enhance any of these aspects of the programming system, the expert must step out of the language, and often out of the programming system entirely.

To put it another way, monolithic programming languages enforce a fundamental dichotomy between \emph{first-class constructs}, which enjoy support throughout the programming system but must be designed and implemented by the system designer in advance, and \emph{user-defined constructs}, which can be distributed as user libraries, but must creatively combine and repurpose the small set of available first-class constructs to express all desired run-time, compile-time and edit-time behaviors. This dichotomy has forced researchers and domain experts who want to significantly advance the state-of-the-art in programming systems, particularly in compiled, statically-typed systems where this dichotomy is most salient, to design a new language in many cases. But this then couples their core innovations with a collection of other, unrelated design choices, requires more effort both for the experts developing the innovation and its targetted user community, and leads to the intrinsic interoperability problems described above for users developing applications consisting of components written in different languages.

\begin{figure}
\begin{center}
\includegraphics[scale=0.5]{approaches.pdf}
\end{center}
\vspace{-20px}
\caption{\small (a) With the language-oriented approach, novel constructs are packaged into separate languages. Users can only safely and naturally call into languages consisting of common constructs. (b) With the library-oriented approach, there is one language and novel constructs are packaged as normal libraries. Thus, interoperability is not a problem but in monolithic systems, this approach is less expressive.\label{approaches}}
\end{figure}

### 2013-01-09 thesis-proposal/timeline.tex
Initial thesis proposal taken from NSF grant

\section{Research Plan}

Our plan for carrying out the work described above involves building significantly on existing work on Ace, as well as iterating on existing highly-simplified prototypes we have developed of all of the components described above.

In the first year, we will fully develop the theory of active types and completing development of the Ace language. We plan on releasing Ace to the broader community, as described in Section 5.1, within the coming year. In addition, we will implement initial versions of a core active code prediction system for Java.

In the second year, we will begin conducting capstone evaluations of Ace, further applying the theory to functional languages, and begin working on applications of active code prediction to code completion and rehabilitative programming systems.

In the third year, we will complete capstone evaluations of Ace, conduct capstone evaluations of active code prediction and the applications that we have developed, and begin prototyping an actively-typed structured editor, laying foundations for the long-term evolution of this research program beyond the scope of this grant.

Evaluations will consist of user studies, case studies, classroom studies and field studies based on instrumented versions of the tools that we develop. We plan on justifying not just the theoretical utility of active types, but their usefulness and usability in practice as well.

### 2013-01-09 thesis-proposal/work.tex
Initial thesis proposal taken from NSF grant

\section{Work To Be Done}
\subsection{Completion and Evaluation of Ace}
The work done so far on the Ace language establishes the practicality of active typechecking and translation as a foundational language mechanism and begins to apply it to one realistic problem domain, high-performance scientific computing on GPUs and other coprocessors, by internalizing OpenCL. However, to more fully establish the practicality of this approach, it must be shown that Ace can support more sophisticated, higher-level language constructs as well, ideally from a variety of application domains. It must also be further shown that users who are not language design experts, and will not be utilizing the extension mechanism directly, can nevertheless use the language in practice.

Thus, a portion of the resources allocated to this project will be dedicated to completing the implementation of Ace, documenting it and releasing it both to the research community and to practitioners in areas where a useful set of constructs have been developed, beginning with the members of the high-performance computing community. This will also support synergy with ongoing collaborations in our group with other problem domains, including front-end and back-end web development, security, functional programming and object-oriented programming.

With broader usage by the research community and in practice, we will be able to continue to conduct case studies and empirical evaluations examining the usability and usefulness of Ace and the active typing mechanisms it introduces. Indeed, because of the flexibility of active typing and our implementation strategy, we plan on instrumenting the Ace compiler to send data and direct feedback from willing users back to us for analysis. This will allow us to analyze questions like:
\begin{itemize}
\item How often are different constructs and mechanisms used in practice?
\item What kinds of errors appear most frequently in practice, and how long does it take users to resolve these errors?
\item How do users feel about the mechanisms, constructs and error messages that they encounter, as determined by direct feedback solicited at the time of occurrence.
\item How commonly do users utilize the active typing mechanism directly?
\end{itemize}

To our knowledge, no such detailed study of usage patterns of a programming language in the wild has been conducted previously, and we anticipate producing insights that are relevant to programming language design broadly.
\subsection{Active Type Theory}
The work on Ace is aimed at establishing a practical implementation of active typechecking and translation. However, because it uses a dynamically-typed language, Python, for the type-level language, it is not suitable for rigorous theoretical analysis, nor does it achieve safety in the strictest sense, due to high levels of dynamism exploitable in Python's object model. Moreover, it is a full language and so it does not necessarily distill the essential concepts that we wish to introduce to the research community in their simplest possible form or connect them to existing concepts in the theory of typed programming languages. For these reasons, we plan to develop a minimal type theoretic formulation, which we call $\lambda_{\text{A}}$.

\subsubsection{Type-Level Computation}

We have chosen to use the term \emph{type-level language}, rather than \emph{metalanguage} or, as it is called in the original work on active libraries, \emph{metacode} intentionally. This is because the concept of a type-level language is already well-established and includes a key feature necessary for active types -- the reification of types as compile-time entities that can be manipulated programmatically. We believe that this connection between type-level computation and active types will provide new insights into each mechanism individually, and provide a clean theoretical basis for active typechecking and translation as an extension to type-level computation.

\subsubsection{Active Types and Wadler's Expression Problem}
Another key insight is that in monolithic programming languages, the fixed set of base types can be written as an algebraic datatype. Indeed, in functional implementations of typecheckers, this is precisely what is done, for example for G\"{o}del's \textbf{T}:

\begin{verbatim}
    datatype Type = Arrow of Type * Type 
                  | Nat
\end{verbatim}

If we wish to allow users to introduce new base types and type families, it can then be seen that we face a variant of the expression problem, so named by Wadler \cite{wadler1998expression}. That is, we wish to add new constructors of kind \verb|Type|, but this conventional formulation of datatypes does not allow us to do so. There are competing approaches that aim to solve this problem. The most common is to use object-oriented inheritance, where subtypes of \verb|Type| represent new type families. This is the approach taken in Ace. The functional programming approach is to instead use \emph{open data types} \cite{conf/ppdp/LohH06}. We will thus aim to further explore this approach, drawing a clear connection between active types, the expression problem and type-level open data types.

\subsubsection{Rigorous Safety Proofs}
A benefit of a simple core theory for active typing is that it will allow us to formally state and prove key safety theorems, formalizing the somewhat informal notions described in Section 3.3. Based on early sketches of a theory that we have discussed, we have implemented some run-time checks into Ace that ensure some of these properties for compiled programs based on a notion borrowed from the area of verified compilation called \emph{type-directed compilation} \cite{tarditi+:til-OLD}. We would like to be able to ensure these properties for all possible programs, and a type theory will allow us to demonstrate the machinery needed to do this in a rigorous manner.

\subsubsection{Implementation With a Dependently-Typed Language}
Finally, we would like to implement a functional version of active type-checking and translation, both as a counterpoint to Ace and as a useful tool for theoretical researchers. We plan on doing so as an extension to the Coq theorem prover, which contains a dependently-typed functional programming language at its core, and is implemented in Ocaml. This will allow us to understand the challenges and opportunities of using a dependently-typed functional language at the core of an extensible programming system.

\subsection{Active Code Prediction}
\subsubsection{Structured, Statistical Code Prediction}

Programming languages are formal systems with rich syntactic and semantic structure. They are
also human systems, in that they are used extensively by people in patterned ways to express their
intent. Many tools are designed to help people write code more efficiently by predicting
the source code that a developer intends. For example, code completion systems for editors like Eclipse for Java display pop-up menus containing the members of the class of the variable being manipulated by the developer.

Typical code completion systems use only this sort of semantic
information about the language and the libraries being used, and do not incorporate data about how developers have written programs
in the past. Recent work by Hindle et al.~\cite{Hindle:2012:NS:2337223.2337322} demonstrated, however, that source code could be successfully predicted statistically
using a simple $n$-gram model that used a tokenized, rather than structural, representation of source code, trained on existing codebases.

Our first aim is to unite the structured and statistical approaches to source code prediction. That is, rather than using a tokenized representation of source code, we would like to do statistical prediction on a more natural representation of source code: the typed syntax tree.  We can then condition our predictions using semantic information, specifically:
\begin{itemize}
\item the \emph{type}, denoted $\tau$, of the expression being predicted (e.g. \verb|int| or \verb|Color|)
\item the \emph{syntactic context}, denoted $\sigma$, in which the expression occurs (e.g. whether the expression is an argument of a function call, the guard of an \verb|if| statement, etc.)
\item the \emph{program context}, denoted $\Gamma$, in which the expression occurs (i.e. the set of variables paired with their types that are in scope at the location that the expression occurs.)
\end{itemize}

For example, if a user has entered the code \texttt{Planet destination = }, where \verb|Planet| is an enumeration containing \verb|Mercury|, \verb|Venus|, \verb|Earth|, etc. (but not \verb|Pluto|, of course), then we have that the expected type at the cursor is \verb|Planet|, the syntactic context is \emph{assignment}, and given a program context, our prediction space should only assign non-zero probabilities to:
\begin{itemize}
\item literal members of the \verb|Planet| enumeration (e.g. \verb|Planet.EARTH|)
\item variables and fields of type \verb|Planet| available from the program context
\item calls to methods available from the program context that have return type \texttt{Planet}\footnote{We can consider  operators like $+$ and $[]$ as methods of the built-in types in Java.}
\end{itemize}
To assign a particular probability to an expression, denoted $e$, we first determine how likely it is that the expression is of each of these three \emph{syntactic forms}, where syntactic forms are denoted $\phi$. For each form, we can then assign probabilities to particular expressions of that form according to some form-specific conditional distribution. This model can be understood as a graphical model, as shown in Figure \ref{graphicalmodel}, where the syntactic form is a latent variable. The conditional distributions for both the syntactic form and expression are learned using data gathered from analyses of prior code corpuses (smoothed using a suitable method in cases where enough information is not available).
\begin{figure}
    \begin{displaymath}
    \xymatrix{
      *+[F]{\color{ForestGreen}\txt{Type ($\tau$)}} \ar[dr] \ar[drr] & & \\
      & *+[F]{\color{black} \txt{Syntactic\\Form ($\phi$)}} \ar[r] & *+[F]{\color{cyan}\txt{Expression ($e$)}}  \\
      *+[F]{\color{ForestGreen}\txt{Syntactic\\Context ($\sigma$)}} \ar[ur] \ar[urr] & & *+[F]{\color{ForestGreen}\txt{Program\\Context ($\Gamma$)}} \ar[u]  
    }
    \end{displaymath}
\caption{\small A graphical model representing our approach. Green variables are always observed (we do not assign marginal distributions to them.) The syntactic form, $\phi$, is a latent variable, and the expression, $e$, is unknown. Note that the syntactic form is a function of the expression. \label{graphicalmodel}}
\end{figure}

\subsubsection{Active Code Prediction}
Given a system for predicting expressions in the context of a type syntax tree, we can then naturally ask: what about specialized domain knowledge that cannot be readily extracted from prior code analysis? For example, consider a type representing strings in a particular language, such as English. Constructors of the \verb|EnglishString| type should be biased to accept literals that have high likelihoods according to some natural language model, but we do not wish to encode all possible natural language models in the core of our prediction system. We can instead take an active types approach by allowing users to equip a type, such as \verb|EnglishString|, with a function, $f_{\text{prob}}$, that takes on the job of predicting the probabilities of only expressions of that type. That is, given a reified representation of $\sigma$, $\Gamma$ and $\phi$, it should be able to produce $P(e | \tau, \phi, \sigma, \Gamma)$. We can then use this prediction directly as if it had been produced by a method described above.

\subsubsection{Foundations, Implementation and Analysis}
We propose developing the theoretical foundations of this idea, in terms of probability theory and type theory, equipping it with a notion of active code prediction as described above, and implementing this atop a real language, likely Java. We have sketched and prototyped a small version of this as an Eclipse extension in the context of a prior course project. We will then evaluate the accuracy of these models, develop use cases as in the related work on active code completion described in Section \ref{graphite}, and analyze the usability and usefulness of this prediction system in applied contexts. In particular, we are interested in uses within code editors to predict expressions and portions of expressions, as well as uses within programming systems for disabled and paralyzed individuals, who need predictive text entry to be able to communicate at a reasonable pace. The information theoretic foundations for this latter application, in the context of brain-computer interfaces, have been developed as prior work by one of the investigators on this proposal \cite{journals/ijhci/OmarAJBMMMC11}.

\subsection{Syntax and Visual Representation}
Thus far, we have been discussing active types within the context of languages with a fixed syntax. However, in many specialized domains, a specialized syntax or visual representation for novel constructs may be appropriate. There are several approaches that we can take to include extensible syntax support in an actively-typed programming system. We have not settled on any one of these, so our proposed research will include a rigorous examination of the pros and cons of each approach.

\subsubsection{Extensible Parsing} 
Extensible parsers have been well-studied in the literature (e.g. \cite{journals/entcs/BrabrandSV03}), and such systems could be integrated with an actively-typed language without conflict with one caveat: each new syntactic form needs to have a dispatch protocol associated with it, to assign control over its semantics to one of its subexpressions. Such a mechanism will be one avenue we will evaluate for suitability.

\subsubsection{Flexible Reinterpretation}
An approach that we have prototyped, but not extensively utilized, in Ace is to allow active types to directly observe the syntax trees of subexpressions, rather than simply their types as we have described above. This allows a highly contextualized reinterpretation of a syntax tree, similar to capabilities found in macro systems (although without a direct rewriting component.) We will further explore this approach, particularly as we begin implementing abstractions that may not naturally fit within Python's existing syntactic forms in an intuitive way.

\subsubsection{Delimited Parsing}
A related alternative approach would be to partially parse a file, leaving delimited regions either unparsed or only tokenized and dispatching to an active type to do the actual parsing. This would give more flexibility to the active type, but also a greater responsibility to fully parse and give semantics to the provided source code. Nevertheless, this may be appropriate in certain cases where a highly-specialized local syntax is desirable for convenience (and flexible reinterpretation or active code completion at edit-time does not suffice.)

\subsubsection{Actively-Typed Structured Editors}
Projects such as Citrus \cite{Ko05} and Barista \cite{Ko06} are analogs of language frameworks for structured editors -- editors where source code elements are represented directly as a tree in a visual or semi-visual form, rather than written as a string as in conventional languages. These tools demonstrate that modular construction of such editors may be possible. An actively-typed structured editor would instead allow an active type to directly control the visual representation and interaction behaviors associated with expressions of that type. This is the most direct analog of active types in this domain, but requires a somewhat radical shift in program representation. It is unlikely that this can be integrated easily into existing languages and systems, as we have largely done in our other work. Thus, we will develop a separate prototype of this approach, and believe that the long-term evolution of actively-typed programming systems will be toward such a representation.

### 2013-01-09 thesis-proposal/listing4.py

@@ -1,0 +1,15 @@
+from ace.OpenCL import OpenCL, get_global_id, 
+  global_ptr, double, int
+
+@OpenCL.fn
+def map(input, output, f):
+    gid = get_global_id(0)
+    output[gid] = f(input[gid])
+
+@OpenCL.fn
+def add5(x):
+    return x + 5
+
+A = GlobalPtrType(double); B = GlobalPtrType(int)
+map_add5_dbl = map.compile(A, A, add5.ace_type)
+map_add5_int = map.compile(B, B, add5.ace_type)

### 2013-01-09 thesis-proposal/listing9.py

@@ -1,0 +1,21 @@
+import ace, ace.astx as astx
+
+class GlobalPtrType(ace.Type):
+  def __init__(self, target_type):
+    self.target_type = target_type
+        
+  def resolve_Subscript(self, context, node):
+    slice_type = context.resolve(node.slice)
+    if isinstance(slice_type, IntegerType):
+      return self.target_type
+    else:
+      raise TypeError('<error message>', node)
+       
+  def translate_Subscript(self, context, node):
+    value = context.translate(node.value)
+    slice = context.translate(node.slice)
+    return astx.copy_node(node, 
+      value=value, slice=slice,    
+      code=value.code + '[' + slice.code + ']')
+      
+  # ...

### 2013-01-10 thesis-proposal/problem.tex
Added updated thesis statement and edited intro a bit

@@
-\section{Problem}
+\section{Introduction}
 
-Designing and implementing a programming language and its supporting tools (collectively, a \emph{programming system}) that simultaneously supports a natural programming style, helps users find and fix errors early, and performs well across diverse problem domains remains a grand challenge in computing.
-Indeed, it is now widely acknowledged that no simple set of constructs will be able to handle all of the challenges modern computing faces.
-Consistent with this view, a large number of researchers and domain experts in both research and industry are actively engaged in the design, implementation and analysis of a variety of novel language-based abstractions, compilation strategies and tools targeted at different problem domains.
+Jonathan Aldrich (ISR), chair
+Brad A. Myers (HCII)
+Robert Harper (CSD)
+Premkumar Devanbu (UCSD)
+
+Designing and implementing a programming language together with its supporting tools (collectively, a \emph{programming system}) that simultaneously supports a natural programming style, helps users find and fix errors early, and consistently performs well across diverse application domains remains a grand challenge in computing.
+Indeed, these concerns are often at odds with one another, so system designers must make trade-offs.
+No small set of general-purpose constructs can completely satisfy every need.
+Evidence supporting this view comes from the large community of researchers and domain experts designing new domain-specific languages and tools, each tailored to the needs of a particular developer community or application domain.
 
 %In implementing a new feature or abstraction, these experts can take one of several approaches. 
-To realize a new feature or abstraction, these experts often take a \emph{language-oriented approach}, whereby the new constructs are developed and distributed together with a new programming language.
-The flexibility of this approach, coupled with support from tools like compiler generators, has led to a proliferation of specialized languages, each built around a particular set of privileged constructs.
+To realize a new construct, these experts often take a \emph{language-oriented approach}, developing and distributing it together with a new programming language.
+The flexibility of this approach, coupled with support from tools like compiler generators and language workbenches, has led to a proliferation of specialized languages, each built around a particular set of privileged constructs.
 Unfortunately, a key issue limiting the practicality of this approach occurs at the boundaries between languages -- specialized constructs particular to one language cannot always be safely and naturally expressed in another, so building a program out of components written in different specialized languages can often prove difficult.

### 2013-03-13 oopsla13/att-oopsla13.tex
Initial port of att paper to oopsla13

%\documentclass[12pt]{article}
\documentclass[10pt]{sigplanconf}
% The following \documentclass options may be useful:
%
% 10pt          To set in 10-point type instead of 9-point.
% 11pt          To set in 11-point type instead of 9-point.
% authoryear    To obtain author/year citation style instead of numeric.
\usepackage{listings}
\usepackage[usenames,dvipsnames]{color}
\usepackage[usenames,dvipsnames]{xcolor}
\usepackage{mathpartir}
\usepackage{amsmath}
\usepackage{amssymb} 
\usepackage{amsthm}
\usepackage{ stmaryrd }
\usepackage{graphicx}

\usepackage[noindentafter]{titlesec}
%\titlespacing{\section}{0pt}{*0.5}{*0.25}
%\titlespacing{\subsection}{3pt}{*1.5}{*0.25}
%\titlespacing{\subsubsection}{3pt}{*1.5}{*0.25}
%\titlespacing{\paragraph}{0pt}{*0.5}{*1}
\usepackage{textpos}
\input{macros}
\input{macros-lamalpha}
\newtheorem{theorem}{Theorem}%[document]
\newtheorem{lemma}{Lemma}
\newtheorem*{corollary}{Corollary}
\newtheorem*{definition}{Definition}

\begin{document}

%\conferenceinfo{POPL '13}{somewhere} 
%\copyrightyear{2013} 
%\copyrightdata{[to be supplied]}

%\titlebanner{{\tt \textcolor{Red}{{\small Under Review -- distribute within CMU only.}}}}        % These are ignored unless
%\preprintfooter{Distribute within CMU only.}   % 'preprint' option specified.

\title{Active Type-Checking and Translation}

%\authorinfo{~}{~}{~}
\authorinfo{Cyrus Omar\and Nathan Fulton\and Jonathan Aldrich}
       {School of Computer Science\\
        Carnegie Mellon University}
        {\{comar, jonathan.aldrich\}@cs.cmu.edu}

\maketitle
\begin{abstract}
% Active Code Completion
% Active Typechecking and Translation
% Active Type-Directed Source Code Prediction
% Actively-Typed Parsing

Researchers and domain experts typically describe new language-based abstractions as extensions of some existing language, but most statically-typed languages are actually \emph{monolithic}: they do not give users the ability to specify the semantics of new core types and operators from within. This choice has led to a proliferation of mutually incompatible standalone languages, each built around a small collection of privileged constructs. 
An alternative to this \emph{language-oriented} approach is to work in an extensible programming language where the compile-time behaviors determining the functionality of new core constructs are specified and implemented modularly within user libraries. Designing such a \emph{library-oriented} extensibility mechanism for statically-typed languages that is both safe and expressive is non-trivial. % -- extensions must not weaken the global safety properties of the language, compromise the integrity of the compilation process, or interfere with one other, while remaining flexible enough to admit a range of useful constructs.
This paper introduces a mechanism called active type-checking and translation (AT\&T) that aims to address these  challenges. By relying upon type-level computation in a novel way, AT\&T admits user specification of a wide range of compile-time behaviors over a fixed grammar in a safe and modular manner. We discuss two points in the design space: (1) a simple calculus designed to distill the essential concepts and admit formal safety theorems, and (2) a fully-implemented language called Ace that we use to demonstrate the expressive power of AT\&T in several real-world domains, including scientific computing, security, functional programming and object-oriented programming.
\end{abstract}

\category{D.3.2}{Programming Languages}{Language Classifications}[Extensible Languages]
\category{D.3.4}{Programming Languages}{Processors}[Compilers]
\category{F.3.1}{Logics \& Meanings of Programs}{Specifying and Verifying and Reasoning about Programs}[Specification Techniques]
%\keywords
%extensible languages, type-level computation, typed compilation

\section{Introduction}
Programming languages have historically been specified and implemented monolithically. To introduce new primitive constructs, researchers or domain experts have developed a new language or a dialect of an existing language, with the help of tools like domain-specific language frameworks and compiler generators \cite{fowler2010domain}. 
Unfortunately, taking a so-called {\it language-oriented approach} \cite{journals/stp/Ward94}, where different languages are used for different components of an application, can lead to problems at language boundaries: a library's external interface must only rely on  constructs that can be expressed in all possible calling languages. This means that specialized invariants cannot be checked statically, decreasing reliability and performance. It also often requires that developers generate verbose and unnatural ``glue'' code, defeating a primary purpose of specialized languages: hiding these low-level details from end-user developers.

Extensible programming languages promise to decrease the need for new standalone languages by providing more granular, language-based support for introducing new primitive constructs (that is, constructs that cannot be adequately expressed in terms of existing syntactic forms and primitive operations.) Developers would gain the freedom to choose those constructs that are most appropriate for their application domain and development discipline. Researchers would gain the ability to distribute new constructs for evaluation by a broader development community without requiring the approval of maintainers of mainstream languages, who are naturally risk-averse and uninterested in niche domains.

A significant challenge that faces language extensibility mechanisms is in maintaining the overall safety properties of the language and compilation process in the presence of arbitrary combinations of user extensions. The mechanism must ensure that basic metatheoretic and global safety guarantees of the language cannot be weakened, that extensions are safely composable, and that type checking and compilation remains decidable. The correctness of an extension itself should be modularly verifiable, so that its users can rely on it for verifying and compiling their own code. These are the issues that we seek to address in this work.

The approach we describe, {\it active type-checking and compilation} (AT\&T), makes use of type-level computation in a novel way. To review, in languages supporting type-level computation, the syntactic class of types is not simply declarative. Instead, it forms a programming language itself (the {\it type-level language}). Types themselves are one kind of value in this language, but there can be many others. To ensure the safety of type-level computations, {\it kinds} classify type-level terms, just as types classify expression-level terms. The simplest example of a language featuring type-level computation is Girard's System $\text{F}_{\omega}$ \cite{tapl}. In $\text{F}_{\omega}$, types have kind $\star$ and type-level functions have kinds like $\star\rightarrow\star$. A growing number of implemented languages now feature more sophisticated type-level languages (see Section 5).  
We emphasize that type-level computation occurs during compilation, rather than at run-time, because type-level terms that are used where types would normally be expected must be reduced to normal form before type-checking can proceed.

In this work, we wish to allow extensions to strengthen the static semantics of our language. Naturally, extension specifications will also need to be evaluated during compilation and manipulate representations of types. This observation suggests that the type-level language may be able to serve directly as a specification language. In this paper, we show that this is indeed the case. By introducing some new constructs at the type-level, developers can specify the semantics of operators associated with newly-introduced families of primitive types with type-level functions. The compiler front-end invokes these functions to synthesize types for and assign meanings to expressions, by translation into a {\it typed internal language}. Unlike conventional metaprogramming systems, these {\it type-level specifications} do not directly manipulate or rewrite expressions. Instead, they examine and manipulate the types of these expressions. 
By using a sufficiently constraining kind system and incorporating techniques from typed compilation into the type-level language directly, the global safety properties of the language and compilation process can be guaranteed. In other words, users can only {\it increase} the safety of the language.

We focus on extending the static semantics of a language with a fixed, though flexible, grammar. Techniques for extensible parsing have been proposed in the past (e.g. \cite{journals/entcs/BrabrandSV03}), and we conjecture that these can be made compatible with the mechanism described in this paper with some simple modifications, but we do not discuss this further here. We also focus on {\it functional}, rather than declarative, specifications of language constructs. Extracting a compiler from a declarative language specification (e.g. in Twelf \cite{twelf}) has not yet been shown practical, but we note that a future mechanism of this sort could safely target a language implementing the mechanism we discuss here.

The organization and key contributions of this paper are:
\begin{itemize}
\item In Section 2, we develop a core calculus, $\lamA$, and give simple examples of language features that can be expressed with it. We formalize the compiler front-end and state several lemmas that lead to useful safety theorems for the compiler and language as a whole.  We show how AT\&T requires that the language provide a solution to a type-level variant of Wadler's expression problem \ref{Wadler}. 
\item In Section 3, we briefly introduce the Ace programming language, which is based fundamentally on an  elaboration of AT\&T that supports a richer set of syntactic forms and a variant of type inference. It uses object-oriented inheritance to solve the type-level expression problem. A number of practical extensions have been written using Ace, including a complete implementation of the OpenCL type system (based on C99) as a library.
\item In Section 4, we briefly describe another point in the design space, a language design we call Birdie. Birdie lifts an extension of the Gallina language, used by the Coq proof assistant, into the type level (leading to a language with dependent kinds). This additional complexity allows for full proofs of correctness for type-level specifications, and can allow proofs soundness of functional specifications against conventional inductive specifications. The expression problem is solved using a constrained formulation of open data types, rather than using object-oriented inheritance. 
\item In Section 5 we compare AT\&T to previous work on extensible languages and compilers, metaprogramming systems and formal specification languages and conclude in Section 6 with a discussion of future work.
\end{itemize}

\section{Type-Level Specifications in $\lamA$}
\subsection{Example: Natural Numbers in $\lamA$} % G\"{o}del's $T$ 
We begin with a simple calculus with no primitive notion of natural numbers, nor any more general notion of an inductive data type. We can, however, concretely specify both the static and dynamic semantics of natural numbers, including the natural recursor of G\"{o}del's $T$ \cite{tapl}, using type-level specifications. Let us begin in Figure 1 with the type $\tvar{nat}$ and its constructors, $\tvar{z}$ and $\tvar{s}$.
\begin{figure}
$\tfam{NAT}{\kunit}{
	\tlam{\tof{\tvar{self}}{\kType{\fvar{NAT}}}}{\titype{\dint}}}{(\\
	~~~~
	\topp{\tops{z}{\tlam{\tof{\tvar{self}}{\kTypeBlur}}{\tconst{\tden{\titerm{0}}{\tvar{self}}}}}}{
	\\~~~~
	\tops{s}{\tlam{\tof{\tvar{self}}{\kTypeBlur}}{\tOp{
		\tlam{\tof{\tvar{d1}}{\kDen}}{
		\\~~~~~~~~~\tifeq{\ttypeof{\tvar{d1}}}{\tvar{self}}{\tden{\tvalof{\tvar{d1}}}{\tvar{self}}}{\terr}}}}}}
	)}{\Theta}{{\textsf{let~}}{\text{nat}=\ttype{\tunit}{\fvar{nat}}}{\textsf{~in~}}{x}}$
\caption{Specifying natural numbers in $\lamA$. This term is referred to as $\psit{nat}$.}
\end{figure}
\begin{figure*}
\small
$$\begin{array}{rccl}
\textbf{expressions} 				&	e	&	::=	&	\evar{x} \pipe 
														\elam{\evar{x}}{\psi}{e} \pipe 
														\eapp{e_{1}}{e_{2}} \pipe
												\eop{\psitype}{op}{
															\splat{\psi}{1}{m}
														}{

														} \pipe
														\eop{\psitype}{op}{
															\splat{\psi}{1}{m}
														}{
  												    		\splat{e}{1}{n}
														} \\
									& 		&		& 	\\

\textbf{type-level specifications} 	& \psi 	& ::= 	& 	\tvar{t} \pipe 
														\tlam{\splatTwo{\tvar{t}}{{:}\kappa}{1}{n}}{\psi} \pipe 
														\tapp{\psi}{\splat{\psi}{1}{n}} \pipe
														\tifeq{\psi_{0}}{\psi_{1}}{\psi_{2}}{\psi_{3}} \\

\text{standard terms}	 			& 		& \pipe	& 	\tunit \pipe 
														\tpair{\psi_{1}}{\psi_{2}} \pipe 
														\tfst{\psi} \pipe 
														\tsnd{\psi} 
														\\

\text{type families}				&		& \pipe	& 	\tfam{fam}{\kappaidx}{\psirep}{\theta}{\Theta}{\psi}\pipe
														\tfamcase{\psi}{\phi}{\psi_0}{\psi_1}\\

\text{types} 						& 		& \pipe	& 	\ttypestd \pipe 
														\tidx{\psitype} \pipe
														\trepof{\psitype}
														\\

\text{operator definitions} 		& 		& \pipe & 	\tconst{\psi} \pipe 
														\tOp{\psi} \\

\text{denotations} 				& 		 & 	\pipe	&	\tden{\psiiterm}{\psitype} \pipe 
														\tvalof{\psiden} \pipe
														\ttypeof{\psiden} \pipe
														\terr \\

\text{internal language}		&		&	\pipe	&	\titerm{\mu} \pipe \titype{\delta}\\

\text{programs}					&		&	\pipe	&	\tprog{e}\\

\text{family specifications}	&	\phi		&	::= & \fvar{fam} \pipe \tfamSpecStd\\
\text{operator lists}		&	\theta	&	::= &	\topsempty \pipe 
												\topp{\theta}{\tops{op}{\psi}}\\

\textbf{kinds} 					& \kappa	&	::=	&	\karrow{\splat{\kappa}{1}{n}}{\kappa} \pipe 
												\kunit \pipe 
												\kpair{\kappa_{1}}{\kappa_{2}} \pipe 
												\kType{\eta} \pipe 
												\kTypeBlur \pipe
												\kOp{n} \pipe
												\kDenk{\eta} \pipe 
												\kDen \pipe 
												\kIType \pipe \kITerm \pipe
												\kProg 
												\\

\text{family signatures}				& \eta		& 	::=	& \kFamStd\\
\text{operator list signatures}			& \Theta	&	::=	&	\cdot \pipe \Theta; \tvar{op}: \kappa\\
											 							&		&		&	\\
\textbf{internal terms} 				& 	\mu	&	::=	&	\ivar{u} \pipe 
												\ilam{\ivar{u}}{\delta}{\mu} \pipe 
												\ifix{\ivar{f}}{\delta}{\mu} \pipe 
												\iapp{\mu_{1}}{\mu_{2}} \pipe 
												\iIfEq{\mu_{0}}{\mu_{1}}{\mu_{2}}{\mu_{3}}\\
							& 		& 	\pipe	&	\iintlit \pipe 
												\iop{\mu_{1}}{\mu_{2}} \pipe 
												\ipair{\mu_{1}}{\mu_{2}} \pipe 
												\ifst{\mu} \pipe
												\isnd{\mu} \pipe
												\iup{\psi} 
												\\
\textbf{internal types}			&	\delta	&	::=	&    \darrow{\delta_1}{\delta_2} \pipe
												\dint \pipe
												\dpair{\delta_1}{\delta_2} \pipe
												\dup{\psi}\\
%\text{cases}					&\tcasev 	&	::=	&	\caseZ{\splat{\kappa}{1}{n}}{\psi} \pipe 
%												\caseS{\tcasev}{\tcasev}\\
%							&		&		&	\\
\end{array}$$
\caption{Syntax of $\lamA$. $\mathbb{Z}$ denotes integer literals and $\klabel{label}$ denotes any label.}
\end{figure*}
\begin{figure*}
\begin{mathpar}
\inferrule{
	\overbrace{\tentailst{\cdot}{\psiprog}{\kProg}}^{\text{\small kind checking}}{ }\\
	\overbrace{\tevalsX{\psiprog}{\tprog{e}}{\noprob}}^{\text{\small specification normalization}}{ }\\
	\overbrace{\translates{\cdot}{e}{\mu}{\ttypestd}}^{\text{\small verification \& translation}}{ }
}{
%	\compiles{\psiprog}{\mu}{\psiidx}{\kappa}{\psirep}
	\compiles{\psiprog}{\mu}{\ttypestd}
}
\end{mathpar}
\caption{Central compilation judgement of $\lamA$.}
\end{figure*}
%\noindent
%
%The \textsf{def} construct is the type-level analog of the standard \textsf{let} construct. Indeed, the entire term consists of type-level constructs. The definition of $\tvar{nat}$ says that it is a type indexed by a uniquely labeled unit (that is, it is a simple type) that will be represented in the internal language by integers. The constructor $\tvar{z}$ is declared as a constant, denoted by the value $0$ and assigned the type $\tvar{nat}$. The successor operator $\tvar{s}$ first checks its argument to ensure that it is a $\tvar{nat}$ (which requires first checking the index's kind to ensure that it has the correct label using the $\sf{idxcase}_{D}$ statement). If so, it returns a denotation that assigns the type $\tvar{nat}$ to an internal term that increments the value of the argument. Otherwise, the special denotation $\terr$ is returned. We will return to explain these constructs in more detail shortly.
%
%This term, which we will call $\psit{nat}$, is an open term because it has not been used to write a program yet. To use this specification to write a program, we substitute a \textsf{program} term containing an expression referring to these definitions for $\tvar{x}$. For example, the following program corresponds to the  number $1$: 
$$\psi_{1} := \subst{\tprog{
	\eop{\text{nat}}{s}{}{\eop{\text{nat}}{z}{}{}}
}}{\psi_{\text{nat}}}$$
%\noindent
%This program will be compiled to the internal term $1 + 0$ and assigned the type $\tvar{nat}$. This notion is formalized by the {\it central compilation judgement} of $\lamA$, given in Figure 3. In this case, we will be able to derive that
$$
\compiless{\psi_{1}}{1 + 0}{\tvar{nat}}
$$
%Compilation will not succeed for invalid terms, e.g.
%\begin{eqnarray*}
%\psi_{-} & := & \subst{\tprog{\econst{\tvar{s}}}}{\tvar{x}}{\psi_{\text{nat}}}\\
%\psi_{-}' & := & \subst{\tprog{\eop{\tvar{s}}{\elam{\evar{n}}{\tvar{nat}}{\eop{\tvar{s}}{\evar{n}}}}}}{\tvar{x}}{\psi_{\text{nat}}}
%\end{eqnarray*}
%Indeed, compilation with $\psi_{\text{nat}}$ as given above admits all well-typed expressions of the simply-typed lambda calculus with natural numbers and no others, compiling $\tvar{nat}$s to internal terms that evaluate to their integer representations.
%
%\subsection{$\lamA$}
\begin{figure*}
$\fbox{\inferrule{}{\tentailsXt{\psi}{\kappa}}}$~
~~~~$\fCtx ::= \cdot \pipe \fCtx, \fSpecStd$
~~~~$\tvarCtx ::= \cdot \pipe \tvarCtx, \tof{\tvar{t}}{\kappa}$
\begin{mathpar}
\small
\inferrule{
	\tof{\tvar{t}}{\kappa} \in \tvarCtx
}{
  \tentailsXt{\tvar{t}}{\kappa}
}~(\text{var}_\psi)

\inferrule{ }{
	\tentailsXt{\tunit}{\kunit}
}~(\text{unit}_\psi)

\inferrule{
	\tentailsXt{\psi_{1}}{\kappa_{1}}\\
	\tentailsXt{\psi_{2}}{\kappa_{2}}
}{
	\tentailsXt{\tpair{\psi_{1}}{\psi_{2}}}{\kpair{\kappa_{1}}{\kappa_{2}}}
}~(\times_\psi)

\inferrule{
	\tentailsXt{\psi}{\kpair{\kappa_{1}}{\kappa_{2}}}
}{
	\tentailsXt{\tfst{\psi}}{\kappa_{1}}
}~(\text{fst}_\psi)

\inferrule{
	\tentailsXt{\psi}{\kpair{\kappa_{1}}{\kappa_{2}}}
}{
	\tentailsXt{\tsnd{\psi}}{\kappa_{2}}
}~(\text{snd}_\psi)

\inferrule{
  \kentailsX{\kappa_1}\\
  \cdots\\
  \kentailsX{\kappa_n}\\
  \tentailst{\tCtxXT{\splatTwo{\tvar{t}}{{:}\kappa}{1}{n}}}{\psi}{\kappa}
}{
  \tentailsXt{\tlam{\splatTwo{\tvar{t}}{{:}\kappa}{1}{n}}{\psi}}{\karrow{\splat{\kappa}{1}{n}}{\kappa}}
}~\text{($\lambda_\psi$)}

\inferrule{
  \tentailsXt{\psi}{\karrow{\splat{\kappa}{1}{n}}{\kappa}}\\
  \tentailsXt{\psi_{1}}{\kappa_{1}}\\
  \cdots\\
  \tentailsXt{\psi_{n}}{\kappa_{n}}
}{
  \tentailsXt{\tapp{\psi}{\splat{\psi}{1}{n}}}{\kappa}
}~(\text{app}_\psi)

\inferrule{
	\tentailsXt{\psi_{0}}{\kappa_{1}}\\
	\tentailsXt{\psi_{1}}{\kappa_{1}}\\
	\tentailsXt{\psi_{2}}{\kappa_{2}}\\
	\tentailsXt{\psi_{3}}{\kappa_{2}}
}{
	\tentailsXt{\tifeq{\psi_{0}}{\psi_{1}}{\psi_{2}}{\psi_{3}}}{\kappa_{2}}
}~(\text{cond}_\psi)

\inferrule{
	\tentailst{\tCtxXF{\fSpecStd}}{\tfamSpecStd}{\kFamStd}\\
	\tentailst{\tCtxXF{\tof{\fvar{fam}}{\kFamStd}}}{\psi}{\kappa}
}{
	\tentailsXt{\tfamStd}{\kappa}
}~(\text{fambind})

\inferrule{
	\tentailsXt{\psitype}{\kTypeBlur}\\
	\tentailsXt{\phi}{\eta}\\
	\tentailsXt{\psi_0}{\karrow{\kType{\eta}}{\kappa}}\\
	\tentailsXt{\psi_1}{\kappa}
}{
	\tentailsXt{\tfamcase{\psitype}{\phi}{\psi_0}{\psi_1}}{\kappa}
}~(\text{famcase-T})

\inferrule{
	\tentailsXt{\psiden}{\kDen}\\
	\tentailsXt{\phi}{\eta}\\
	\tentailsXt{\psi_0}{\karrow{\kDenk{\eta}}{\kappa}}\\
	\tentailsXt{\psi_1}{\kappa}
}{
	\tentailsXt{\tfamcase{\psiden}{\phi}{\psi_0}{\psi_1}}{\kappa}
}~(\text{famcase-D})

\inferrule{
\tentailsXt{\phi}{\kFamStd}\\
\tentailsXt{\psiidx}{\kappaidx}
}{
	\tentailsXt{\ttype{\psiidx}{\phi}}{\kType{\kFamStd}}
} ~(\text{Type}_I)

\inferrule{
	\tentailsXt{\psi}{\kType{\eta}}
}{
	\tentailsXt{\psi}{\kTypeBlur}
}~(\text{Type-$\subseteq$})

\inferrule{
	\tentailsXt{\psi}{\kType{\kFamStd}}
}{
	\tentailsXt{\tidx{\psi}}{\kappaidx}
}~(\text{idxof})

\inferrule{
	\tentailsXt{\psi}{\kTypeBlur}
}{
	\tentailsXt{\trepof{\psi}}{\kIType}
}~(\text{repof})

\inferrule{
	\tentailsXt{\psi}{\kDen}
}{
	\tentailsXt{\tconst{\psi}}{\kOp{0}}
}~(\text{const})

\inferrule{
	\tentailsXt{\psi}{\karrow{\kDen, \stackrel{n}{\cdots}, \kDen}{\kDen}}
}{
	\tentailsXt{\tOp{\psi}}{\kOp{n}}
}~(\text{$n$-op})

\inferrule{
	\tentailsXt{\psiiterm}{\kITerm}\\
	\tentailsXt{\psitype}{\kType{\eta}}
}{
	\tentailsXt{\tden{\psiiterm}{\psitype}}{\kDenk{\eta}}
}~(\text{Den}_I)

\inferrule{
	\tentailsXt{\psiiterm}{\kITerm}\\
	\tentailsXt{\psitype}{\kTypeBlur}
}{
	\tentailsXt{\tden{\psiiterm}{\psitype}}{\kDen}
}~(\text{Den}_I\text{-$\subseteq$})

\inferrule{
	\tentailsXt{\psi}{\kDenk{\eta}}
}{
	\tentailsXt{\psi}{\kDen}
}~(\text{Den-$\subseteq$})

\inferrule{ }{
	\tentailsXt{\terr}{\kDen}
}~(\text{err})

\inferrule{
	\tentailsXt{\psi}{\kDenk{\eta}}
}{
	\tentailsXt{\ttypeof{\psi}}{\kType{\eta}}
}~(\text{typeof})

\inferrule{
	\tentailsXt{\psi}{\kDenk{\eta}}
} {
	\tentailsXt{\tvalof{\psi}}{\kITerm}
}~(\text{valof})
\\
\inferrule{
	\ientails{\iMkCtx{\fCtx}{\tvarCtx}{\cdot}}{\mu}
}{
	\tentailsXt{\titerm{\mu}}{\kITerm}	
}~(\text{iterm})

\inferrule{
	\tentailsX{\delta}
}{
	\tentailsXt{\titype{\delta}}{\kIType}
}~(\text{itype})

\inferrule{
	\ientails{\iMkCtx{\fCtx}{\tvarCtx}{\cdot}}{e}
}{
	\tentailsXt{\tprog{e}}{\kProg}
}~(\text{program})
\end{mathpar}
$\fbox{\inferrule{}{\tentailsXt{\phi}{\eta}}}$~
\begin{mathpar}
\small
\inferrule{
	\tof{\fvar{fam}}{\eta} \in \Sigma
}{
	\tentailsXt{\fvar{fam}}{\eta}
}~(\text{var}_\phi)

\inferrule{
	\kentailsX{\kFamStd}\\
	\tentailsXt{\psirep}{\karrow{\kappaidx}{\kIType}}\\
	\tentailsXt{\theta}{\Theta}
}{
	\tentailsXt{\tfamSpecStd}{\kFamStd}
}~(\text{family}_I)
\end{mathpar}
$\fbox{\inferrule{}{\tentailsXt{\theta}{\Theta}}}$
\begin{mathpar}
\small
\inferrule{ }{
	\tentailsXt{\cdot}{\cdot}
}~(\text{no-ops})

\inferrule{
	\tentailst{\Sigma, \tof{\fvar{fam}}{\eta}~\Delta}{\theta}{\Theta}\\
	\tentailst{\Sigma, \tof{\fvar{fam}}{\eta}~\Delta}{\psi}{\karrow{\kType{\eta}}{\kappa}}
}{
	\tentailst{\Sigma, \tof{\fvar{fam}}{\eta}~\Delta}{\topp{\theta}{\tops{id}{\psi}}}{\Topp{\Theta}{\Tops{id}{\kappa}}}
}~(\text{op-seq})
\end{mathpar}
\caption{Kind checking rules for $\lamA$. All contexts have standard substructural properties (not shown).}
\end{figure*}
\begin{figure*}
$\fbox{\inferrule{}{\kentailsX{\kappa}}}$~
$\fbox{\inferrule{}{\kentailsX{\kSimple{\kappa}}}}$
%~~~~$\$
\begin{mathpar}
\small
\inferrule{
	\kentailsX{\kappa_1}\\
	\cdots\\
	\kentailsX{\kappa_n}\\
	\kentailsX{\kappa}
}{
	\kentailsX{\karrow{\splat{\kappa}{1}{n}}{\kappa}}
}~(\text{$\rightarrow$}_\kappa)

\inferrule{
	\kentailsX{\kappa_1}\\
	\kentailsX{\kappa_2}
}{
	\kentailsX{\kpair{\kappa_1}{\kappa_2}}
}~(\text{$\times$}_\kappa)

\inferrule{
	\kentailsX{\kSimple{\kappa}}
}{
	\kentailsX{\kappa}
}~(\text{simple})

\inferrule{
	\kentailsX{\kSimple{\kappa_1}}\\
	\kentailsX{\kSimple{\kappa_2}}
}{
	\kentailsX{\kSimple{\kpair{\kappa_1}{\kappa_2}}}
}~(\text{$\times_\kappa$-simple})

\inferrule{ }{
	\kentailsX{\kSimple{\kunit}}
}~(\text{Unit}_\kappa)

\inferrule{ }{
	\kentailsX{\kSimple{\kTypeBlur}}
}~(\text{Type-$\subseteq$}_\kappa)

\inferrule{
	\kentailsX{\eta}
}{
	\kentailsX{\kSimple{\kType{\eta}}}
}~(\text{Type}_\kappa)

\inferrule{ }{
	\kentailsX{\kSimple{\kDen}}
}~(\text{Den-$\subseteq$}_\kappa)

\inferrule{
	\kentailsX{\eta}
}{
	\kentailsX{\kSimple{\kDenk{\eta}}}
}~(\text{Den}_\kappa)

\inferrule{ }{
	\kentailsX{\kSimple{\kIType}}
}~(\text{IType}_\kappa)

\inferrule{ }{
	\kentailsX{\kSimple{\kITerm}}
}~(\text{ITerm}_\kappa)

\inferrule{ }{
	\kentailsX{\kSimple{\kProg}}
}~(\text{Program}_\kappa)

\inferrule{ }{
	\kentailsX{\kSimple{\kOp{n}}}
}~(\text{$n$-Op}_\kappa)
\end{mathpar}
$\fbox{\inferrule{}{\kentailsX{\eta}}}$~
\begin{mathpar}
\small
\inferrule{
	\kentailsX{\kSimple{\kappaidx}}\\
	\kentailsX{\Theta}
}{
	\kentailsX{\kFamStd}
}~(\text{family}_\eta)
\end{mathpar}
$\fbox{\inferrule{}{\kentailsX{\Theta}}}$
\begin{mathpar}
\small
\inferrule{~}{
	\kentailsX{\cdot}
}~(\text{No-Ops})
%
%\inferrule{
%	\kentailsX{\Theta}
%}{
%	\kentailsX{\Topp{\Theta}{\Tops{id}{\kOp{n}}}}
%}~(\text{(0, $n$)-Op-Seq})

\inferrule{
	\kentailsX{\Theta}\\
	\kentailsX{\karrow{\splat{\kappa}{1}{m}}}{\kOp{n}}
}{
	\kentailsX{\Topp{\Theta}{\Tops{id}{\karrow{\splat{\kappa}{1}{m}}{\kOp{n}}}}}
}~(\text{($m$, $n$)-Op-Seq})
\end{mathpar}
$\fbox{\inferrule{}{\eentailsX{e}}}$
~~~~$\etvarCtx ::= \cdot \pipe \etvarCtx, \evar{x}$
\begin{mathpar}
\small
\inferrule{
	\evar{x} \in \etvarCtx
}{
	\eentailsX{\evar{x}}
}~(\text{var-form}_e)

\inferrule{
	\tentailsXt{\psi}{\kTypeBlur}\\
	\eentails{\fCtx}{\tvarCtx}{\etvarCtx, \evar{x}}{e}
}{
	\eentailsX{\elam{\evar{x}}{\psi}{e}}
}~(\text{$\lambda$-form}_e)

\inferrule{
	\eentailsX{e_1}\\
	\eentailsX{e_2}
}{
	\eentailsX{\eapp{e_1}{e_2}}
}~(\text{app-form}_e)

\inferrule{
	\tentailsXt{\psitype}{\kType{\kFamStd}}\\
	\tof{\tvar{id}}{\karrow{\splat{\kappa}{1}{m}}{\kOp{0}}}\in\Theta\\
	\tentailsXt{\psi_1}{\kappa_1}\\
	\cdots\\
	\tentailsXt{\psi_m}{\kappa_m}
}{
	\eentailsX{\eop{\psitype}{\tvar{id}}{\splat{\psi}{1}{m}}{ }}
}~(\text{$(m, 0)$-op-app-form}_e)

\inferrule{
	\tentailsXt{\psitype}{\kType{\kFamStd}}~~~~~~~~
	\tof{\tvar{id}}{
		\karrow{\splat{\kappa}{1}{m}}{
			\kOp{n}
		}
	} \in \Theta~~~~~~~~
	\tentailsXt{\psi_1}{\kappa_1}~~~~~~~~
	\cdots~~~~~~~~
	\tentailsXt{\psi_m}{\kappa_m}\\
	\eentailsX{e_1}\\
	\cdots\\
	\eentailsX{e_n}
}{
	\eentailsX{\eop{\psitype}{\tvar{id}}{\splat{\psi}{1}{m}}{\splat{e}{1}{n}}}
}~(\text{$(m, n)$-op-app-form}_e)
\end{mathpar}
$\fbox{\inferrule{}{\ientailsX{\mu}}}$
~~~~$\itvarCtx ::= \cdot \pipe \itvarCtx, \ivar{u}$
\begin{mathpar}
\small
\inferrule{
	\ivar{u} \in \itvarCtx
}{
	\ientailsX{\ivar{u}}
}~(\text{var-form}_{\mu})

\inferrule{
	\tentailsX{\delta}\\
	\ientails{\iMkCtx{\fCtx}{\tvarCtx}{\itvarCtx}, \ivar{u}}{\mu}
}{
	\ientailsX{\ilam{\ivar{u}}{\delta}{\mu}}
}~(\text{$\lambda$-form}_{\mu})

\inferrule{
	\tentailsX{\delta}\\
	\ientails{\iMkCtx{\fCtx}{\tvarCtx}{\itvarCtx}, \ivar{f}}{\mu}
}{
	\ientailsX{\ifix}{\ivar{f}}{\delta}{\mu}
}~(\text{fix-form}_\mu)

\inferrule{
	\ientailsX{\mu_1}\\
	\ientailsX{\mu_2}
}{
	\ientailsX{\iapp}{\mu_1}{\mu_2}
}~(\text{app-form}_\mu)

\inferrule{
	\ientailsX{\mu_1}\\
	\ientailsX{\mu_2}\\
	\ientailsX{\mu_3}\\
	\ientailsX{\mu_4}
}{
	\ientailsX{\iIfEq}{\mu_1}{\mu_2}{\mu_3}{\mu_4}
}~(\text{cond-form}_\mu)

\inferrule{ }{\ientailsX{\iintlit}}~(\text{int-form}_\mu)

\inferrule{
	\ientailsX{\mu_1}\\
	\ientailsX{\mu_2}
}{
	\ientailsX{\iop{\mu_1}{\mu_2}}
}~(\text{+-form}_\mu)

\inferrule{
	\ientailsX{\mu_1}\\
	\ientailsX{\mu_2}
}{
	\ientailsX{\ipair{\mu_1}{\mu_2}}
}~(\text{$\times$-form}_\mu)

\inferrule{
	\ientailsX{\mu}
}{
	\ientailsX{\ifst{\mu}}
}~(\text{fst-form}_\mu)

\inferrule{
	\ientailsX{\mu}
}{
	\ientailsX{\isnd{\mu}}
}~(\text{snd-form}_\mu)

\inferrule{
	\tentailsXt{\psi}{\kITerm}
}{
	\ientailsX{\iup{\psi}}
}~(\text{$\uparrow$-form}_\mu)
\end{mathpar}
$\fbox{\inferrule{}{\tentailsX{\delta}}}$
\begin{mathpar}
\small
\inferrule{
	\tentailsX{\delta_1}\\
	\tentailsX{\delta_2}
}{
	\tentailsX{\darrow{\delta_1}{\delta_2}}
}~(\text{$\rightarrow$-form}_\delta)

\inferrule{ }{\tentailsX{\dint}}~(\text{int-form}_\delta)

\inferrule{
	\tentailsX{\delta_1}\\
	\tentailsX{\delta_2}
}{
	\tentailsX{\dpair{\delta_1}{\delta_2}}
}~(\text{$\times$-form}_\delta)

\inferrule{
	\tentailsXt{\psi}{\kIType}
}{
	\tentailsX{\dup{\psi}}
}~(\text{$\uparrow$-form}_\delta)
\end{mathpar}
\caption{Kind and term formation rules}
\end{figure*}

\begin{figure*}
%  \renewcommand\thefigure{4}
% % \begin{textblock}{20}(0,0)
%  \newpage
$\fbox{$\tevalX{\psi}{\psi'}$}$~
$\fbox{$\tevalX{\theta}{\theta'}$}$
$\fbox{$\tevalX{\phi}{\phi'}$}$
~~~~$\famEvalCtx ::= \cdot \pipe \famEvalCtx,\tfamSpecStd$
~~~~~~~~$\errCtx ::= \noprob \pipe \prob \pipe \errCtx, \errCtx$
\begin{mathpar}
\small
\inferrule{ }{
	\tevalsX{\tunit}{\tunit}{\noprob}
}~(\text{unit}_\beta)

\inferrule{
	\tevalsX{\psi_1}{\psi_1'}{\errCtx_{1}}\\
	\tevalsX{\psi_2}{\psi_2'}{\errCtx_{2}}
}{
	\tevalsX{\tpair{\psi_1}{\psi_2}}{\tpair{\psi_1'}{\psi_2'}}{\errCtx_{1},\errCtx_{2}}
}~(\text{pair}_\beta)

\inferrule{
	\tevalX{\psi}{\tpair{\psi_1}{\psi_2}}
}{
	\tevalX{\tfst{\psi}}{\psi_1}
}~(\text{fst}_\beta)

\inferrule{
	\tevalX{\psi}{\tpair{\psi_1}{\psi_2}}
}{
	\tevalX{\tsnd{\psi}}{\psi_2}
}~(\text{snd}_\beta)

\inferrule{ }{
	\tevalsX{\tlam{\splatTwo{\tvar{t}}{{:}\kappa}{1}{n}}{\psi}}{\tlam{\splatTwo{\tvar{t}}{{:}\kappa}{1}{n}}{\psi}}{\noprob}
}~(\lambda_\beta)

\inferrule{
	\tevalsX{\psi_0}{\tlam{\splatTwo{\tvar{t}}{{:}\kappa}{1}{n}}{\psi}}{\errCtx_{0}}\\
	\tevalsX{\psi_1}{\psi_1'}{\errCtx_{1}}\\
	\cdots\\
	\tevalsX{\psi_n}{\psi_n'}{\errCtx_{n}}\\
	\tevalsX{\substn{\splatTwo{\psi'}{{/}t}{1}{n}}{\psi}}{\psi'}{\errCtx}
}{
	\tevalsX{\tapp{\psi_0}{\splat{\psi}{1}{n}}}{\psi'}{\errCtx_{0},\errCtx_{1},\ldots,\errCtx_{n},\errCtx}
}~(\text{app}_\beta)

\inferrule{
	\tevalsX{\psi_0}{\psi_0'}{\errCtx_{0}}\\
	\tevalsX{\psi_1}{\psi_0'}{\errCtx_{1}}\\
	\tevalsX{\psi_2}{\psi_2'}{\errCtx_{2}}
}{
	\tevalsX{\tifeq{\psi_0}{\psi_1}{\psi_2}{\psi_3}}{\psi_2'}{\errCtx_{0},\errCtx_{1},\errCtx_{2}}
}~(\text{if}_\beta^1)

\inferrule{
	\tevalsX{\psi_0}{\psi_0'}{\errCtx_{0}}\\
	\tevalsX{\psi_1}{\psi_1'}{\errCtx_{1}}\\
	\psi_0' \neq \psi_1'\\
	\tevalsX{\psi_3}{\psi_3'}{\errCtx_{3}}
}{
	\tevalsX{\tifeq{\psi_0}{\psi_1}{\psi_2}{\psi_3}}{\psi_3'}{\errCtx_{0},\errCtx_{1},\errCtx_{3}}
}~(\text{if}_\beta^2)

\inferrule{
	\tevalsX{\tfamSpecStd}{\phi}{\errCtx_0}\\
	\tevalsX{\subst{\phi}{\fvar{fam}}{\psi}}{\psi'}{\errCtx_1}
}{
	\tevalsX{\tfamStd}{\psi'}{\errCtx_0, \errCtx_1}
}~(\text{fambind}_\beta)

\inferrule{
	\tevalsX{\psirep}{\psirep'}{\errCtx_0}\\
	\tevalsX{\theta}{\theta'}{\errCtx_1}
}{
	\tevalsX{\tfamSpecStd}{\tfamSpec{fam}{\kappaidx}{\psirep'}{\theta'}{\Theta}}{\errCtx_0, \errCtx_1}
}~(\text{family}_\beta)

\inferrule{
	\tevalsX{\psitype}{\ttype{\psiidx}{\phi'}}{\errCtx_0}\\
	\tevalsX{\phi}{\phi'}{\errCtx_1}\\
	\tevalsX{\tapp{\psi_0}{\ttype{\psiidx}{\phi'}}}{\psi_0'}{\errCtx_2}
}{
	\tevalsX{
		\tfamcase{\psitype}{\phi}{\psi_0}{\psi_1}
	}{
		\psi_0'
	}{\errCtx_0, \errCtx_1, \errCtx_2}
}~(\text{famcase-T}_\beta^1)

\inferrule{
	\tevalsX{\psitype}{\ttype{\psiidx}{\phi'}}{\errCtx_0}\\
	\tevalsX{\phi}{\phi''}{\errCtx_1}\\
	\phi' \neq \phi''\\
	\tevalsX{\psi_1}{\psi_1'}{\errCtx_2}
}{
	\tevalsX{
		\tfamcase{\psitype}{\phi}{\psi_0}{\psi_1}
	}{
		\psi_1'
	}{\errCtx_0, \errCtx_1, \errCtx_2}
}~(\text{famcase-T}_\beta^2)

\inferrule{
	\tevalsX{\psiden}{\tden{\psiiterm}{\ttype{\psiidx}{\phi'}}}{\errCtx_0}\\
	\tevalsX{\phi}{\phi'}{\errCtx_1}\\
	\tevalsX{\tapp{\psi_0}{\tden{\psiiterm}{\ttype{\psiidx}{\phi'}}}}{\psi_0'}{\errCtx_2}
}{
	\tevalsX{
		\tfamcase{\psiden}{\phi}{\psi_0}{\psi_1}
	}{
		\psi_0'
	}{\errCtx_0, \errCtx_1, \errCtx_2}
}~(\text{famcase-D}_\beta^1)

\inferrule{
	\tevalsX{\psiden}{\tden{\psiiterm}{\ttype{\psiidx}{\phi'}}}{\errCtx_0}\\
	\tevalsX{\phi}{\phi''}{\errCtx_1}\\
	\phi' \neq \phi''\\
	\tevalsX{\psi_1}{\psi_1'}{\errCtx_2}
}{
	\tevalsX{
		\tfamcase{\psiden}{\phi}{\psi_0}{\psi_1}
	}{
		\psi_1'
	}{\errCtx_0, \errCtx_1, \errCtx_2}
}~(\text{famcase-D}_\beta^2)

\inferrule{
	\tevalsX{\psiidx}{\psiidx'}{\errCtx}
}{
	\tevalsX{\ttypestd}{
		\ttype{\psiidx'}{fam}
	}{\errCtx}
}

\inferrule{
	\tevalsX{\psitype}{\ttypestd}{\errCtx}
}{
	\tevalsX{\tidx{\psitype}}{\psiidx}{\errCtx}
}

\inferrule{
	\tevalsX{\psitype}{\ttypestd}{\errCtx_0}\\
	\tfamSpecStd \in \famEvalCtx\\
	\tevalsX{\tapp{\psirep}{\psiidx}}{\psi'}{\errCtx_1}
}{
	\tevalsX{\trepof{\psitype}}{\psi'}{\errCtx_0, \errCtx_1}
}

\inferrule{
	\tevalsX{\psi}{\psi'}{\errCtx}
}{
	\tevalsX{\tconst{\psi}}{\tconst{\psi'}}{\errCtx}
}

\inferrule{
	\tevalsX{\psi}{\psi'}{\errCtx}
}{
	\tevalsX{\tOp{\psi}}{\tOp{\psi'}}{\errCtx}
}

\inferrule{
	\tevalsX{\psiiterm}{\psiiterm'}{\errCtx_0}\\
	\tevalsX{\psitype}{\psitype'}{\errCtx_1}
}{
	\tevalsX{\tden{\psiiterm}{\psitype}}{\tden{\psiiterm'}{\psitype'}}{\errCtx_0, \errCtx_1}
}

\inferrule{
	\tevalsX{\psiden}{\tden{\psiiterm}{\psitype}}{\errCtx}
}{
	\tevalsX{\tvalof{\psiden}}{\psiiterm}{\errCtx}
}

\inferrule{
	\tevalsX{\psiden}{\tden{\psiiterm}{\psitype}}{\errCtx}
}{
	\tevalsX{\ttypeof{\psiden}}{\psitype}{\errCtx}
}

\inferrule{ }{
	\tevalsX{\terr}{\terr}{\noprob}
}

\inferrule{
	\tevalsX{\mu}{\mu'}{\errCtx}
}{
	\tevalsX{\titerm{\mu}}{\titerm{\mu'}}{\errCtx}
}

\inferrule{
	\tevalsX{\delta}{\delta'}{\errCtx}
}{
	\tevalsX{\titype{\delta}}{\titype{\delta'}}{\errCtx}
}

\inferrule{
	\tevale{e}{e'}
}{
	\tevalX{\tprog{e}}{\tprog{e'}}
}

\inferrule{ }{
	\tevalsX{\topsempty}{\topsempty}{\noprob}
}

\inferrule{
	\tevalsX{\theta}{\theta'}{\errCtx_0}\\
	\tevalsX{\psi}{\psi'}{\errCtx_1}
}{
	\tevalsX{
		\topp{\theta}{\tops{op}{\psi}}
	}{
		\topp{\theta'}{\tops{op}{\psi'}}
	}{\errCtx_0, \errCtx_1}
}
\end{mathpar}
$\fbox{$\tevale{e}{e'}$}$
\begin{mathpar}
\small
\inferrule{
	\tevalsX{\psi}{\psi'}{\errCtx}
}{
	\tevalsX{\elam{\evar{x}}{\psi}{e}}
	{\elam{\evar{x}}{\psi'}{e}}{\errCtx}
}

\inferrule{
	\tevalsX{e_1}{e_1'}{\errCtx_1}\\
	\tevalsX{e_2}{e_2'}{\errCtx_2}
}{
	\tevalsX{\eapp{e_1}{e_2}}{\eapp{e_1'}{e_2'}}{\errCtx_1, \errCtx_2}
}

\inferrule{
	\tevalsX{\psitype}{\psitype'}{\errCtx}\\
	\tevalsX{\psi_1}{\psi_1'}{\errCtx_1^1}\\
	\cdots\\
	\tevalsX{\psi_m}{\psi_m'}{\errCtx_m^1}\\
	\tevalsX{e_1}{e_1'}{\errCtx_1^2}\\
	\cdots\\
	\tevalsX{e_n}{e_n'}{\errCtx_n^2}
}{
	\tevalsX{\eop{\psitype}{op}{\splat{\psi}{1}{m}}{\splat{e}{1}{n}}}
	{\eop{\psitype'}{op}{\splat{\psi'}{1}{m}}{\splat{e'}{1}{n}}}
	{\errCtx, \errCtx_1^1, \ldots, \errCtx_m^1, \errCtx_1^2, \ldots, \errCtx_n^2}
}
\end{mathpar}
$\fbox{$\tevalm{\mu}{\mu'}$}$
\begin{mathpar}
\small
\inferrule{
	\tevalsX{\delta}{\delta'}{\errCtx}
}{
	\tevalsX{\ilam{\ivar{u}}{\delta}{\mu}}{\ilam{\ivar{u}}{\delta'}{\mu}}{\errCtx}
}

\inferrule{
	\tevalsX{\delta}{\delta'}{\errCtx}
}{
	\tevalsX{\ifix{\ivar{f}}{\delta}{\mu}}{\ifix{\ivar{f}}{\delta'}{\mu}}{\errCtx}
}

\inferrule{
	\tevalsX{\mu_1}{\mu_1'}{\errCtx_1}\\
	\tevalsX{\mu_2}{\mu_2'}{\errCtx_2}
}{
	\tevalsX{\iapp{\mu_1}{\mu_2}}{\iapp{\mu_1'}{\mu_2'}}{\errCtx_1, \errCtx_2}
}

\inferrule{
	\tevalsX{\mu_0}{\mu_0'}{\errCtx_0}\\
	\tevalsX{\mu_1}{\mu_1'}{\errCtx_1}\\
	\tevalsX{\mu_2}{\mu_2'}{\errCtx_2}\\
	\tevalsX{\mu_3}{\mu_3'}{\errCtx_3}
}{
	\tevalsX{\iIfEq{\mu_0}{\mu_1}{\mu_2}{\mu_3}}{\iIfEq{\mu_0'}{\mu_1'}{\mu_2'}{\mu_3'}}{\errCtx_0, \errCtx_1, \errCtx_2, \errCtx_3}
}

\inferrule{ }{
	\tevalsX{\iintlit}{\iintlit}{\noprob}
}

\inferrule{
	\tevalsX{\mu_1}{\mu_1'}{\errCtx_1}\\
	\tevalsX{\mu_2}{\mu_2'}{\errCtx_2}
}{
	\tevalsX{\iop{\mu_1}{\mu_2}}{\iop{\mu_1'}{\mu_2'}}{\errCtx_1, \errCtx_2}
}

\inferrule{
	\tevalsX{\mu_1}{\mu_1'}{\errCtx_1}\\
	\tevalsX{\mu_2}{\mu_2'}{\errCtx_2}
}{
	\tevalsX{\ipair{\mu_1}{\mu_2}}{\ipair{\mu_1'}{\mu_2'}}{\errCtx_1, \errCtx_2}
}

\inferrule{
	\tevalsX{\mu}{\mu'}{\errCtx}
}{
	\tevalsX{\ifst{\mu}}{\ifst{\mu'}}{\errCtx}
}

\inferrule{
	\tevalsX{\mu}{\mu'}{\errCtx}
}{
	\tevalsX{\isnd{\mu}}{\isnd{\mu'}}{\errCtx}
}

\inferrule{
	\tevalsX{\psi}{\titerm{\mu}}{\errCtx}
}{
	\tevalsX{\iup{\psi}}{\mu}{\errCtx}
}
\end{mathpar}
$\fbox{$\tevalm{\delta}{\delta'}$}$
\begin{mathpar}
\small
\inferrule{
	\tevalsX{\delta_1}{\delta_1'}{\errCtx_1}\\
	\tevalsX{\delta_2}{\delta_2'}{\errCtx_2}
}{
	\tevalsX{\darrow{\delta_1}{\delta_2}}{\darrow{\delta_1'}{\delta_2'}}{\errCtx_1, \errCtx_2}
}

\inferrule{ }{
	\tevalsX{\dint}{\dint}{\noprob}
}

\inferrule{
	\tevalsX{\delta_1}{\delta_1'}{\errCtx_1}\\
	\tevalsX{\delta_2}{\delta_2'}{\errCtx_2}
}{
	\tevalsX{\dpair{\delta_1}{\delta_2}}{\dpair{\delta_1'}{\delta_2'}}{\errCtx_1, \errCtx_2}
}

\inferrule{
	\tevalsX{\psi}{\titype{\delta}}{\errCtx}
}{
	\tevalsX{\dup{\psi}}{\delta}{\errCtx}
}
\end{mathpar}
%
%\inferrule{
%	\tevalX{\psi}{\ttype{\psiidx}{\kappa}{\psirep}}
%}{
%	\tevalX{\tidx{\psi}}{\psiidx}
%}
%
%\inferrule{
%	\tevalX{\psi}{\ttype{\psiidx}{\kappa}{\psirep}}
%}{
%	\tevalX{\trepof{\psi}}{\psirep}
%}
%
%\inferrule{
%	\tevalsX{\psi}{\ttype{\psiidx}{\kappa}{\psirep}}{\errCtx_{0}}\\
%	\tevalsX{\tapp{\psi_{0}}{\ttype{\psiidx}{\kappa}{\psirep}}}{\psi'}{\errCtx_{1}}
%}{
%	\tevalsX{\tfamcase{\psi}{\kappa}{\psi_{0}}{\psi_{1}}}{\psi'}{\errCtx_{0},\errCtx_{1}}
%}
%
%\inferrule{
%	\tevalsX{\psi}{\ttype{\psiidx}{\kappa'}{\psirep}}{\errCtx_{0}}\\
%	\kappa \neq \kappa'\\
%	\tevalsX{\psi_{1}}{\psi'}{\errCtx_{1}}
%}{
%	\tevalsX{\tfamcase{\psi}{\kappa}{\psi_{0}}{\psi_{1}}}{\psi'}{\errCtx_{0},\errCtx_{1}}
%}
%
%\inferrule{
%	\tevalX{\psi}{\psi'}
%}{
%	\tevalX{\tconst{\psi}}{\tconst{\psi'}}
%}
%
%\inferrule{
%	\tevalX{\psi}{\psi'}
%}{
%	\tevalX{\tOp{\psi}}{\tOp{\psi'}}
%}
%
%\inferrule{
%	\tevalms{\mu}{\mu'}{\errCtx_{0}}\\
%	\tevalsX{\psitype}{\ttype{\psiidx}{\kappa}{\psirep}}{\errCtx_{1}}\\
%	\mtentails{\cdot}{\mtof{\mu'}{\psirep}}
%}{
%	\tevalsX{\tden{\mu}{\psitype}}{\tden{\mu'}{\psitype'}}{\errCtx_{0},\errCtx_{1}}
%}
%
%\inferrule{
%	\tevalms{\mu}{\mu'}{\errCtx_{0}}\\
%	\tevalsX{\psitype}{\ttype{\psiidx}{\kappa}{\psirep}}{\errCtx_{1}}\\
%	\cdot \nvdash \mtof{\mu'}{\psirep}
%}{
%	\tevalsX{\tden{\mu}{\psitype}}{\tden{\mu'}{\psitype'}}{\errCtx_{0},\errCtx_{1},\prob}
%}
%
%\inferrule{
%	\tevalX{\psi}{\tden{\mu}{\psitype}}
%}{
%	\tevalX{\ttypeof{\psi}}{\psitype}
%}
%
%\inferrule{ }{
%	\tevalsX{\terr}{\terr}{\noprob}
%}
%
%\inferrule{
%	\tevalsX{\psi}{\tden{\mu}{\ttype{\psiidx}{\kappa}{\psirep}}}{\errCtx_{0}}\\
%	\tevalsX{\tapp{\psi_{0}}{\tden{\mu}{\ttype{\psiidx}{\kappa}{\psirep}}}}{\psi'}{\errCtx_{1}}
%}{
%	\tevalsX{\ttidxcase{\psi}{\kappa}{\psi_{0}}{\psi_{1}}}{\psi'}{\errCtx_{0},\errCtx_{1}}
%}
%
%\inferrule{
%	\tevalsX{\psi}{\terr}{\errCtx_{0}}\\
%	\tevalsX{\psi_{1}}{\psi'}{\errCtx_{1}}
%}{
%	\tevalsX{\ttidxcase{\psi}{\kappa}{\psi_{0}}{\psi_{1}}}{\psi'}{\errCtx_{0},\errCtx_{1}}
%}
%
%\inferrule{
%	\tevalsX{\psi}{\tden{\mu}{\ttype{\psiidx}{\kappa'}{\psirep}}}{\errCtx_{0}}\\
%	\kappa \neq \kappa'\\
%	\tevalsX{\psi_{1}}{\psi'}{\errCtx_{1}}
%}{
%	\tevalsX{\ttidxcase{\psi}{\kappa}{\psi_{0}}{\psi_{1}}}{\psi'}{\errCtx_{0},\errCtx_{1}}
%}
%
%\inferrule{
%	\tevalsX{\psi_1}{\psi_1'}{\errCtx_{0}}\\
%	\tevalsX{\psi_2}{\psi_2'}{\errCtx_{1}}
%}{
%	\tevalsX{\tILtarrow{\psi_1}{\psi_2}}{\tILtarrow{\psi_1'}{\psi_2'}}{\errCtx_{0},\errCtx_{1}}
%}
%
%\inferrule{
%	\tevalsX{\psi_1}{\psi_1'}{\errCtx_{0}}\\
%	\tevalsX{\psi_2}{\psi_2'}{\errCtx_{1}}
%}{
%	\tevalsX{\tILtprod{\psi_1}{\psi_2}}{\tILtprod{\psi_1'}{\psi_2'}}{\errCtx_{0},\errCtx_{1}}
%}
%
%\inferrule{ }{
%	\tevalsX{\tILtint}{\tILtint}{\noprob}
%}
%
%\inferrule{
%	\tevalsX{\psi}{\tILtarrow{\psi_1}{\psi_2}}{\errCtx_{0}}\\
%	\tevalsX{\tapp{\psiarrow}{\psi_1, \psi_2}}{\psi'}{\errCtx_{1}}
%}{
%	\tevalsX{\tILtcase{\psi}{\psiarrow}{\psiprod}{\psiint}}{\psi'}{\errCtx_{0},\errCtx_{1}}
%}
%
%\inferrule{
%	\tevalsX{\psi}{\tILtprod{\psi_1}{\psi_2}}{\errCtx_{0}}\\
%	\tevalsX{\tapp{\psiprod}{\psi_1, \psi_2}}{\psi'}{\errCtx_{1}}
%}{
%	\tevalsX{\tILtcase{\psi}{\psiarrow}{\psiprod}{\psiint}}{\psi'}{\errCtx_{0},\errCtx_{1}}
%}
%
%\inferrule{
%	\tevalsX{\psi}{\tILtint}{\errCtx_{0}}\\
%	\tevalsX{\psiint}{\psiint'}{\errCtx_{1}}
%}{
%	\tevalsX{\tILtcase{\psi}{\psiarrow}{\psiprod}{\psiint}}{\psiint'}{\errCtx_{0},\errCtx_{1}}
%}
%
%\inferrule{ }{
%	\tevalms{\ivar{u}}{\ivar{u}}{\noprob}
%}
%
%\inferrule{
%	\tevalX{\psiden}{\tden{\mu}{\psitype}}
%}{
%	\tevalm{\mvalof{\psiden}}{\mu}
%}
%
%\inferrule{
%	\tevalsX{\psi}{\psi'}{\errCtx_{0}}\\
%	\tevalms{\mu}{\mu'}{\errCtx_{1}}
%}{
%	\tevalms{\ilam{\ivar{u}}{\psi}{\mu}}{\ilam{\ivar{u}}{\psi'}{\mu'}}{\errCtx_{0},\errCtx_{1}}
%}
%
%\inferrule{
%	\tevalsX{\psi}{\psi'}{\errCtx_{0}}\\
%	\tevalms{\mu}{\mu'}{\errCtx_{1}}
%}{
%	\tevalms{\ifix{\ivar{f}}{\psi}{\mu}}{\ifix{\ivar{f}}{\psi'}{\mu'}}{\errCtx_{0},\errCtx_{1}}
%}
%
%\inferrule{
%	\tevalms{\mu_1}{\mu_1'}{\errCtx_{0}}\\
%	\tevalms{\mu_2}{\mu_2'}{\errCtx_{1}}
%}{
%	\tevalms{\iapp{\mu_1}{\mu_2}}{\iapp{\mu_1'}{\mu_2'}}{\errCtx_{0},\errCtx_{1}}
%}
%
%\inferrule{ }{
%	\tevalms{\iintlit}{\iintlit}{\noprob}
%}
%
%\inferrule{
%	\tevalms{\mu_1}{\mu_1'}{\errCtx_{0}}\\
%	\tevalms{\mu_2}{\mu_2'}{\errCtx_{1}}
%}{
%	\tevalms{\iop{\mu_1}{\mu_2}}{\iop{\mu_1'}{\mu_2'}}{\errCtx_{0},\errCtx_{1}}
%}
%
%\inferrule{
%	\tevalms{\mu_1}{\mu_1'}{\errCtx_{0}}\\
%	\tevalms{\mu_2}{\mu_2'}{\errCtx_{1}}
%}{
%	\tevalms{\ipair{\mu_1}{\mu_2}}{\ipair{\mu_1'}{\mu_2'}}{\errCtx_{0},\errCtx_{1}}
%}
%
%\inferrule{
%	\tevalm{\mu}{\mu'}
%}{
%	\tevalm{\ifst{\mu}}{\ifst{\mu'}}
%}
%
%\inferrule{
%	\tevalm{\mu}{\mu'}
%}{
%	\tevalm{\isnd{\mu}}{\isnd{\mu'}}
%}
%
%\inferrule{
%	\tevalms{\mu_0}{\mu_0'}{\errCtx_{0}}\\
%	\tevalms{\mu_1}{\mu_1'}{\errCtx_{1}}\\
%	\tevalms{\mu_2}{\mu_2'}{\errCtx_{2}}\\
%	\tevalms{\mu_3}{\mu_3'}{\errCtx_{3}}
%}{
%	\tevalms{\iIfEq{\mu_0}{\mu_1}{\mu_2}{\mu_3}}{\iIfEq{\mu_0'}{\mu_1'}{\mu_2'}{\mu_3'}}{\errCtx_{0},\errCtx_{1},\errCtx_{2},\errCtx_{3}}
%}
%
%\inferrule{ }{
%	\tevales{\evar{x}}{\evar{x}}{\noprob}
%}
%
%\inferrule{
%	\tevalsX{\psi}{\psi'}{\errCtx_{0}}\\
%	\tevales{e}{e'}{\errCtx_{1}}
%}{
%	\tevales{\elam{\evar{x}}{\psi}{e}}{\elam{\evar{x}}{\psi'}{e'}}{\errCtx_{0},\errCtx_{1}}
%}
%
%\inferrule{
%	\tevales{e_1}{e_1'}{\errCtx_{0}}\\
%	\tevales{e_2}{e_2'}{\errCtx_{1}}
%}{
%	\tevales{\eapp{e_1}{e_2}}{\eapp{e_1'}{e_2'}}{\errCtx_{0},\errCtx_{1}}
%}
%
%\inferrule{
%	\tevalX{\psi}{\psi'}
%}{
%	\tevale{\econst{\psi}}{\econst{\psi'}}
%}
%
%\inferrule{
%	\tevalsX{\psi}{\psi'}{\errCtx_{0}}\\
%	\tevales{e_1}{e_1'}{\errCtx_{1}}\\
%	\cdots\\
%	\tevales{e_n}{e_n'}{\errCtx_{n}}
%}{
%	\tevales{\eop{\psi}{\splat{e}{1}{n}}}{\eop{\psi'}{\splat{e'}{1}{n}}}{\errCtx_{0},\errCtx_{1},\ldots,\errCtx_{n}}
%}
\caption{Big-step operational semantics for type-level terms. The error context $\Sigma$ is used to detect representational consistency errors.}
%%\end{textblock}
\end{figure*}
%\begin{figure*}
%\renewcommand\thefigure{5}
%$\fbox{$\translates{\eCtx}{e}{\mu}{\psiidx}{\kappa}{\psirep}$}$
%~~~~$\eCtx ::= \cdot \pipe \extendeCtx{x}{\mu}{\psiidx}{\kappa}{\psirep}$
%\begin{mathpar}
%\inferrule{
%	\translatesTo{\evar{x}}{\mu}{\psiidx}{\kappa}{\psirep} \in \eCtx
%}{
%	\translatesX{\evar{x}}{\mu}{\psiidx}{\kappa}{\psirep}
%}
%
%\inferrule{ }{
%	\translatesX{
%		\econst{\tconst{\tden{\mu}{\ttype{\psiidx}{\kappa}{\psirep}}}}
%	}{
%		\mu
%	}{
%		\psiidx
%	}{
%		\kappa
%	}{
%		\psirep
%	}
%}
%
%\inferrule{
%	\translates{\extendeCtx{x}{\ivar{x}}{\psiidx}{\kappa}{\psirep}}{e}{\mu'}{\psiidx'}{\kappa'}{\psirep'}
%}{
%	\translatesX{
%		\elam{\evar{x}}{\ttype{\psiidx}{\kappa}{\psirep}}{e}
%	}{
%		\ilam{\ivar{x}}{\psirep}{\mu'}
%	}{
%		\tpair{\ttype{\psiidx}{\kappa}{\psirep}}{\ttype{\psiidx'}{\kappa'}{\psirep'}}
%	}{
%		\kpair{\kType{\kappa}}{\kType{\kappa'}}
%	}{
%		\tILtarrow{\psirep}{\psirep'}
%	}
%}
%
%\inferrule{
%	\translatesX{e_1}{\mu_1}{\psiidxn{1}}{\kappa_1}{\psirepn{1}}~~~~
%	\cdots~~~~
%	\translatesX{e_n}{\mu_n}{\psiidxn{n}}{\kappa_n}{\psirepn{n}}\\
%	\tevalsX{\tapp{\psi}{
%		\tden{\mu_1}{\ttype{\psiidxn{1}}{\kappa_1}{\psirepn{1}}},
%		\ldots,
%		\tden{\mu_n}{\ttype{\psiidxn{n}}{\kappa_n}{\psirepn{n}}}
%	}}{\tden{\mu}{\ttype{\psiidx}{\kappa}{\psirep}}}{\noprob}
%}{
%	\translatesX{
%		\eop{\tOp{\psi}}{\splat{e}{1}{n}}
%	}{
%		\mu
%	}{
%		\psiidx
%	}{
%		\kappa
%	}{
%		\psirep
%	}
%}
%
%\end{mathpar}
%\caption{Verification and translation rules for $\lamA$.}
%\end{figure*}
%\begin{figure*}
%  \renewcommand\thefigure{6}
%$\fbox{$\mtentailsXt{\mu}{\psi}$}$
%~~~~$\iCtx ::= \cdot \pipe \iCtx, \mtof{\ivar{u}}{\psi}$
%\begin{mathpar}
%\inferrule{
%	\mtof{\ivar{u}}{\psi} \in \ivarCtx
%}{
%	\mtentailsXt{\ivar{u}}{\psi}
%}
%
%\inferrule{
%	\mtentails{\iCtx, \mtof{\ivar{u}}{\psi}}{\mtof{\mu}{\psi'}}
%}{
%	\mtentailsXt{\ilam{\ivar{u}}{\psi}{\mu}}{\psi'}
%}
%
%\inferrule{
%	\mtentails{\iCtx, \mtof{\ivar{f}}{\psi}}{\mtof{\mu}{\psi}}
%}{
%	\mtentailsXt{\ifix{\ivar{f}}{\psi}{\mu}}{\psi}
%}
%
%\inferrule{
%	\mtentailsXt{\mu_1}{\tILtarrow{\psi_1}{\psi_2}}\\
%	\mtentailsXt{\mu_2}{\psi_1}
%}{
%	\mtentailsXt{\iapp{\mu_1}{\mu_2}}{\psi_2}
%}
%
%\inferrule{ }{
%	\mtentailsXt{\iintlit}{\tILtint}
%}
%
%\inferrule{
%	\mtentailsXt{\mu_1}{\tILtint}\\
%	\mtentailsXt{\mu_2}{\tILtint}
%}{
%	\mtentailsXt{\iop{\mu_1}{\mu_2}}{\tILtint}
%}
%
%\inferrule{
%	\mtentailsXt{\mu_1}{\psi_1}\\
%	\mtentailsXt{\mu_2}{\psi_2}
%}{
%	\mtentailsXt{\ipair{\mu_1}{\mu_2}}{\tILtprod{\psi_1}{\psi_2}}
%}
%
%\inferrule{
%	\mtentailsXt{\mu}{\tILtprod{\psi_1}{\psi_2}}
%}{
%	\mtentailsXt{\ifst{\mu}}{\psi_1}
%}
%
%\inferrule{
%	\mtentailsXt{\mu}{\tILtprod{\psi_1}{\psi_2}}
%}{
%	\mtentailsXt{\isnd{\mu}}{\psi_2}
%}
%
%\inferrule{
%	\mtentailsXt{\mu_0}{\psi_1}\\
%	\mtentailsXt{\mu_1}{\psi_1}\\
%	\mtentailsXt{\mu_2}{\psi_2}\\
%	\mtentailsXt{\mu_3}{\psi_2}
%}{
%	\mtentailsXt{\iIfEq{\mu_0}{\mu_1}{\mu_2}{\mu_3}}{\psi_2}
%}
%\end{mathpar}
%\caption{The internal type system of $\lamA$.}
%\end{figure*}

%Let us now examine the definition of $\lamA$ in more detail. The grammar of $\lamA$ is given in Figure 2. Expressions in $\lamA$ can take one of only five forms, the first three of which correspond to the simply-typed lambda calculus and the remaining two to abstract constants and operator invocations, respectively. 
%
%The type-level specification language is based on System F \cite{tapl} equipped with several additional primitives that we will describe below.  To emphasize the expanded role for type-level terms, we use the letter $\psi$ instead of the more conventional $\tau$. It is constrained by a kind system, the rules of which are given in Figure 4. A big-step semantics for the type-level language is given in Figure 5. As we will state more formally below, well-kinded terms of the type-level language strongly normalize.
%
%Expressions are given meaning in a denotational style in terms of a typed internal language according to the verification and translation rules in Figure 6. The type system of the internal language is given in Figure 7. Unlike the type-level language, the internal language contains a ${\sf fix}$ operator, and is thus Turing-complete (corresponding to a variant of Plotkin's PCF \cite{tapl}.)
%
%\subsubsection{Central Compilation Judgement}
%The central compilation judgement of Figure 3 takes a type-level term that reduces to a program and produces a typed denotation for the program if and only if it is valid with respect to the kind system of $\lamA$ {\it and} the type system introduced by the user, the key feature of this mechanism. It consists of three steps:
%\begin{enumerate}
%\item \textbf{Kind checking}, where basic correctness properties of the user specification as well as uses of type-level terms in expressions are checked.
%\item \textbf{Specification Normalization}, which reduces the specification to normal form by internalizing relevant portions of it directly into the expressions of the program.
%\item \textbf{Verification and Translation}, where operator definitions are invoked to assign final denotations to expressions if possible, or derive that there is an $\terr$ if not.
%\end{enumerate}
%The rules given in Figures 4-7 are sufficient to define the central compilation judgement. We examine its theoretical properties in Section 2.4.
%
%\subsubsection{Type Constructors}
%
%An interesting feature of $\lamA$ is that there is just one form of type constructor: 
%$\ttype{\psiidx}{\kappa}{\psirep}$.
%All types are identified by their index, $\psiidx$ and an associated internal type, $\psirep$, which controls the representation of values that inhabit the type. The principal kind of a type constructor indexed by a value of kind $\kappa$ is $\kType{\kappa}$. 
%
%

afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf

afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf

afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf

afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf

afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf

afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf

afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf

afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf

afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf

afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf

afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf

afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf

afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf

afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf

afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf

afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf

afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf

afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf

afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf

afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf

afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf

afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf

afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf

afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf

afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf

afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf

%In the example in Figure 1, $\tvar{nat}$ is a type with no (that is, unit) index that we will represent by an integer in the internal language. To distinguish it from other types of the same basic form, we need to label it. That is the purpose of the \textsf{label} declaration preceding the type definition -- it introduces a new kind of unit labeled $\klabel{nat}$. We can then identify the type $\tvar{nat}$ as being the sole type indexed by a value of kind $\klabel{nat}$ and represented by the internal type $\tILtint$. It is important to note that although the type $\tvar{nat}$ has kind $\kType{\klabel{nat}}$, it does not uniquely inhabit that kind -- you can construct other types indexed by a $\klabel{nat}$-labeled unit but associated with different internal types, so operator definitions must test for equality as appropriate.
%
%\subsubsection{Internal Terms and Types}
%The internal language, $\mu$, is a variant of Plotkin's PCF with integers, products and a conditional checks for $\beta$-equality. The form $\mvalof{\psi}$ is used to compose denotations, which are internal language terms paired with a type constructor (see below). In the type-level operational semantics, the $\mvalof{\psi}$ form  always reduces to one of the other forms of the internal language.
%
%Internal types are represented at the type-level by the constructors \texttt{arrow}, \texttt{prod} and \texttt{int}, which have kind $\kIType$. The internal type system is given in Figure 7. The language is equipped with the standard type safety theorem based on preservation and progress lemmas. Because expressions are given their semantics ultimately in terms of internal language terms, these global safety theorems are inherited by the language as a whole. In a more realistic language, additional desirable global safety restrictions could also be imposed using the internal type system (e.g. memory safety).
%
%\subsubsection{Function Types}
%It is worth considering now what type to assign to expression-level functions, e.g. $\elam{\evar{x}}{\tvar{nat}}{\eop{\tvar{s}}{\eop{\tvar{s}}{x}}}$. Because we have only one form for types, we must encode arrow types using this form. The encoding, given formally in Figure 6, encodes the arrow type $\psi_{1} \rightarrow \psi_{2}$, where $\psi_{1,2}$ are types indexed by values of kind $\kappa_{1,2}$, as
%$$\ttype{\tpair{\psi_{1}}{\psi_{2}}}{\kpair{\kType{\kappa_{1}}}{\kType{\kappa_{2}}}}{\tILtarrow{\trepof{\psi_{1}}}{\trepof{\psi_{2}}}}$$
%\noindent
%That is, the arrow type is a type indexed by a pair containing the constituent types and represented by the internal arrow type. As can be seen in the compilation rules, expression-level functions translate directly into internal language functions.
%\subsubsection{Operators}
%In order to construct and manipulate terms of a newly introduced type, we must provide primitive operators, introduced using the forms \textsf{const} and \textsf{op}. The kind associated with operators is $\kOp{n}$, where $n$ is the operator's arity. An operator must have fixed arity. Constants can be thought of as nullary operators. 
%
%Constants are given meaning directly using a denotation. The constant $\tvar{z}$, for example, has the denotation $\tden{0}{\tvar{nat}}$, reflecting the fact that it is a constant of type $\tvar{nat}$ represented by the value $0$. Expressions of the form $\econst{\psi}$ compile to the denotation associated with $\psi$ if $\psi$ is a constant, as can be seen in Figure 6.
%
%Operators of higher arity, $n$, are parameterized by a type-level function that takes $n$ denotations (of kind $\kDen$, see below). This function must analyze the types associated with these denotations, then return a new denotation composed using the input denotations to give meaning to expressions of the form $\eop{\psi}{\splat{e}{1}{n}}$. Note that this implies that operators must have a strictly compositional semantics -- neither expressions nor internal terms can be directly manipulated in the specification language.
%
%Because type checking and evaluation of function application is not compositional (requiring extending a context and substitution, respectively), it cannot be simulated by an operator definition in $\lamA$ as stated.
%
%\subsubsection{Denotations}
%Although operator implementations take arguments of kind $\kDen$, a denotation itself does not have principal kind $\kDen$. Recall that denotations contain a type, and types have kind $\kType{\kappa}$, where $\kappa$ is the kind of the type's index. To support the extraction of a type from a denotation, we must ``remember'' its index kind even when it has been packed into a denotation. As a result, denotations have principal kind $\kDenk{\kappa}$. 
%
%To allow for operators that are polymorphic over a number of types (but not necessarily {\it all} types), we allow for the index kind to be erased as needed. To reintroduce the kind, the $\textsf{idxcase}_{D}$ form is available to perform conditional kind analysis during normalization. This can be thought of as a simplified form of subkinding with casts. Analagous erasure semantics exist for types themselves, with a corresponding {\sf idxcase} form. 
%
%To ensure that this does not allow for non-terminating terms, we impose a strict positivity restriction on type indices that contain values of kind $\kTypeBlur$. This condition, which restricts $\kTypeBlur$ so that it cannot appear to the left of any arrows, is found in many total functional programming languages that feature inductive datatypes, such as Coq \cite{bertot04book} and Agda \cite{norell2007towards}, and we refer the reader to these sources for a detailed discussion.
%
%Another important reason for the distinction between $\kDen$ and $\kDenk{\kappa}$ is that there must be some mechanism by which an operator specification indicates a type error. In $\lamA$, this corresponds to a special value, $\terr$, which is also of kind $\kDen$. However, no type is associated with the denotation $\terr$, so it can never be given the kind $\kDenk{\kappa}$. The verification and translation rules in Figure 6 ensure that $\terr$ is never propagated past the operator that first produces it. 
%
%\subsection{G\"{o}del's $T$ in $\lamA$}
%In Section 2.1, we showed an encoding of natural numbers in $\lamA$, $\psit{nat}$. It would be instructive now for the reader to review the examples in that section and derive the various components of the compilation judgement for them. For the two negative examples, it is interesting to see why they fail to compile -- the first does not pass kind checking while the second is a user-defined error caught by the definition of $\tvar{s}$. 
%
%We can now complete the encoding of G\"{o}del's $T$ in $\lamA$. The missing piece is the natural recursor. For review, the typing rule for the natural recursor can be written (omitting contexts):
%\begin{mathpar}
%\inferrule{
%	{e_{1}:\tvar{nat}}\\
%	{e_{2}:\tau}\\
%	{e_{3}:\tvar{nat}\rightarrow\tau\rightarrow{\tau}}
%}{
%	{\tt natrec}~e_{1}~e_{2}~e_{3} : \tau
%}
%\end{mathpar}
%The corresponding operator definition in $\lamA$, written in a concrete syntax that omits keywords and ``{\sf else err}'' branches is:
%\begin{verbatim}
%def natrec := forall k. op(fun (d0 d1 d2 : Den) := 
%  didxcase d0 of NAT => 
%  if d0 = nat => 
%  didxcase d1 of k => 
%  didxcase d2 of Type[NAT]*(Type[k]*Type[k]) =>
%  def t1 := typeof d1 in 
%  def t2 := typeof d2 in 
%  if fst idx t2 = nat =>
%  def s := snd idx t2 in 
%  if fst s = snd s => 
%  if fst s = t1 => 
%  [[(fix f (n :: int) :: rep t1 :=
%       if n = 0 then valof(d1) 
%       else valof(d2) (n - 1) (f (n - 1))
%     ) valof(d0), t1]]) in x
%\end{verbatim}
%Here, we see that natrec is parameterized by the kind of the index of the type $\tau$ above, corresponding to \texttt{t1} with index kind \texttt{k}. The body of the definition amounts to checking that the equational relations between the types for the three arguments that are implied by the inductive definition actually hold. The implementation of the natural recursor is a standard recursive definition that mirrors the operational semantics usually given to it. If we call this specification $\psit{natrec}$, then the composition of the two specifications produces a specification that is sound and complete against $T$:
%$$\psi_{T} := \subst{\psit{natrec}}{\tvar{x}}{\psit{nat}}$$
%
%One might ask at this point what the benefit is of using this more complex specification style? For simple examples like this, it is clearly more complex than necessary. However, we will see in Section 4 examples where a purely inductive specification of a language construct requires significantly more machinery than a functional specification. In cases where the simpler inductive style is convenient, on the other hand, a sound translation from an inductive specification language to a naive type-level specification would not be difficult to develop. Moreover, automatically extracting an efficient implementation from an inductive specification of a semantics is generally far more difficult. With type-level specification, we err on the side of greater control whenever safe.
%\subsection{Safety of $\lamA$}
%As mentioned in the introduction, evaluating complex type-level functions to verify and translate programs is fraught with theoretical peril. Both the decidability of type-checking and the metatheoretic and global safety properties of the language itself are at risk of disruption by unwise user specifications. Fortunately, the formulation of $\lamA$ allows us to prove two important safety theorems. 
%
%The first guarantees that the compilation judgement of Figure 2, which includes the type checking and translation phases, is decidable.
%\begin{theorem}[Decidability of Compilation]
%There is a terminating decision procedure that can derive the compilation judgement for any term $\psi$ if it exists, or halt if it does not exist.
%\end{theorem}
%\noindent
%The second guarantees that successful compilation results in a well-typed internal language term that is representationally consistent with its type assignment.
%\begin{theorem}[Type-Preserving Compilation]
%If 
%$$\compiles{\psi}{\mu}{\psiidx}{\kappa}{\psirep}$$ 
%then $\mtentails{\cdot}{\mtof{\mu}{\psirep}}$.
%\end{theorem}
%
%\subsubsection{Metatheoretic Properties of the Type-Level Language}
%There are three main lemmas that we must show to ensure that compilation is decidable.
%\begin{lemma}[Decidability of Kind Checking]
%Derivation of the kind assignment judgement $\tentailsXt{\psi}{\kappa}$ is decidable.
%\end{lemma}
%\begin{lemma}[Decidability of Internal Type Checking]
%Derivation of the internal type checking judgement $\mtentailsXt{\mu}{\psi}$ is decidable if $\psi$ is a canonical form for kind $\kIType$.
%\end{lemma}
%\begin{lemma}[Strong Normalization to Canonical Form] If a closed type-level term $\psi$ is well-kinded, that is 
% $\tentails{\cdot}{\cdot}{\cdot}{\tof{\psi}{\kappa}}$,  then $\tevalsX{\psi}{\psi'}{\Sigma}$ for some $\Sigma$ and $\psi'$ is a canonical form of kind $\kappa$.
%\end{lemma}
%\noindent
%We omit statements of the canonical forms lemma and other minor lemmas due to space considerations. Lemma 3 is the most difficult to prove but also absolutely critical -- a type-level language that does not guarantee termination will necessarily lead to undecidable type checking. Our approach, which we do not have space to detail here, is based on the standard logical relations formulation for strong normalization in System F, with special considerations of type indices. The check for strictly positive occurrences of the kind $\kTypeBlur$ ensures that values of kind $\kTypeBlur$ cannot sneak self-references into functions by clever use of {\sf idxcase}, by an argument exactly analagous to those given for total functional programming languages with inductive data types. In fact, we'll see in the next section that this analogy between type constructors and inductive data types has significance in practice.
%
%Using these three lemmas, we can show by simple inductive argument that the verification and compilation rules of Figure 5 are decidable, and thus that Theorem 1 holds.
%\subsubsection{Representational Consistency}
%Theorem 2 can be shown straightforwardly by noting that the big-step evaluation rules for denotations check representational consistency directly during normalization steps. The error context $\Sigma$ tracks whether any denotation constructed during the normalization and verification and translation processes is representationally inconsistent. If so, the final evaluation judgement cannot be derived. Combined with the representational consistency of the rule for arrow types, described in Section 2.2.2, we have a stronger theorem than the one stated -- that no denotation created during the compilation of the program $\psi$ is representationally inconsistent.
%
%The issue is that this form of checking amounts to ``dynamic checking'' of this particular invariant at the type-level. In particular, we cannot check that a well-kinded specification will produce a valid denotation or $\terr$ for all possible programs that it is applied to -- only those branches of the specification that are exercised for a particular program are checked. This means that users will occasionally be faced with specification errors even after the kind checking phase, in addition to any errors arising from problems in their programs.
%
%The reason we can't statically check specifications for this error is that it is a relational property between the two components of a denotation. As such, it requires much more sophisticated support for equational reasoning in the kind system. In $\psit{nat}$, for example, proving that even $\tvar{z}$ uses a representationally consistent denotation requires static knowledge that $\trepof{\tvar{nat}}=\tILtint$. While this is easily discernable from the specification as given, the simply-kinded system given does not track enough information to prove this kind of property. The problem is more clearly evident in the specifications for $\tvar{s}$ and $\tvar{natrec}$, which contain uses of \textsf{valof}. To prove that the internal term $1 + \mvalof{\psi}$ is well-typed without evaluating $\psi$ requires a proof that $\trepof{\ttypeof{\psi}}=\tILtint$. To prove that the denotation is representationally consistent then requires the same proof that $\trepof{\tvar{nat}}=\tILtint$. In Section 3, we give a method for translating type-level terms into Coq, which supports such reasoning. In practice, coverage tests to ensure that all branches of a specification are exercised by tests may be sufficient for prototype implementations of language features, with rigorous equational reasoning reserved for those features being deployed more widely or in safety-critical applications. Indeed, even in certified compilers like CompCert, the type checking and early translation phases are written in a general-purpose functional language (e.g. Ocaml) for practicality \cite{Leroy-Compcert-CACM}.
%
%\subsubsection{Local vs. Global Safety Guarantees}
%While Theorem 2 guarantees that any global properties guaranteed by the internal type system will hold for all compiled programs, this mechanism does not support the introduction of new global safety properties for all possible programs. For instance, although the implementation of System $T$ described here can be shown sound and complete, some of the safety properties guaranteed by the constructs of $\psi_{T}$ hold only locally -- in program fragments that only use operators defined in $\psi_{T}$. A user could easily introduce a new constant $\tvar{b} := \tconst{\tden{-1}{\tvar{nat}}}$, for example. The introduction of this constant invalidates the mapping between the natural numbers of mathematics and the type $\tvar{nat}$, effectively redefining it.
%
%The purpose of this section was to demonstrate a technique that allows for the concrete specification and modular introduction of new primitives into a language. This facilitates experimentation with collections of primitives known (or at least strongly suspected) to be mutually consistent. But just as introducing new axioms into a proof assistant can cause problems if not done carefully, composing primitives should also be done with care. The hope is that by making experimentation as easy as importing a library and ensuring that it cannot invalidate fundamental properties of the language, different communities will more quickly converge upon powerful collections of mutually consistent primitives.
%
%In the next section, we sketch a general technique for proving stronger safety and composability theorems about specifications.
%\section{Strengthening Specifications with Dependent Kinds}
%In the previous section, we gave a minimal calculus that expresses the main ideas of this paper -- a type-level specification language that contains a uniform representation for types and operators. Operators were implemented as type-level functions operating on both indexed types and denotations. Although the language and compilation process could be shown to be safe in the face of these extensions, the extensions themselves could have errors in them -- we discussed representational consistency, soundness with respect to an independent specification (e.g. mathematical natural numbers) and composability of global guarantees.
%
%These issues all have in common that they require making complex statements about specifications. Proof assistants based in type theory such as Coq, Agda, Twelf and others are often used by programming languages researchers for just that task. By taking advantage of dependent types married with flexible inductive datatypes, these proof assistants are emerging as powerful tools for certified programming. They also possess the fundamental properties required of our metalanguage -- strong normalization.
%
%To practically realize a language featuring type-level specifications and demonstrate techniques that could be used for certifying these specifications, we propose lifting one of these languages into the type level. The key issue is that these languages do not admit a simple embedding for uniform type constructors. In most language specifications, types are represented using an inductive data type with a finite set of cases. In $\lamA$ however, types can be introduced anywhere. This is an instantiation of the well-known expression problem of Wadler \cite{wadler1998expression}. 
%
%A number of techniques have been proposed in the literature to resolve this issue in general-purpose functional languages. Open data types are one popular approach \cite{conf/ppdp/LohH06}. The indexed types in $\lamA$ can be considered an open data kind where each possible index kind generates a new constructor. Each constructor must contain a slot to place the associated internal type's representation. We use this observation to construct a translation from a language in the style of $\lamA$ to Gallina, the dependently-typed language of the Coq proof assistant. The kinds that are possible for indices can be read statically from a type-level term. Index case statements containing closed kinds can thus be translated into \texttt{match} statements containing two cases: one for to the corresponding constructor, and one matching all other constructors. Index case statements containing kind variables (e.g. natrec) require a more sophisticated approach, however. The case kind must be unified against all possible closed index kinds and a match case must be generated for each one where unification succeeds. 
%
%All the remaining constructs needed by $\lamA$ can be expressed using standard inductive data types. By also implementing the language and compiler itself in Gallina and providing a specification of the semantics of the internal language, type-level terms can be used to reason about the behavior of operator implementations in nearly arbitrary detail. A proof-passing approach to representational consistency as discussed in Section 2.4.3  can guarantee that users will only ever see errors originating in their own expressions because specifications themselves have been modularly verified.
%
%To prove strong composability results about a collection of types and primitives, there are a number of possible approaches. For the primitives of $\psi_{T}$, for example, it can be seen that it does not interfere with other primitives -- all denotations are assigned either a newly introduced type -- $\tvar{nat}$ -- or an external type -- ${\tt t1}$ -- paired with an internal term taken directly from an input denotation to the operator and also having type {\tt t1}, which we can inductively assume is semantically consistent. Proving such a theorem, while not trivial, is well within the capabilities of Coq and other proof assistants. Our choice of Coq is largely motivated by familiarity as well as Coq's sophisticated program extraction capabilities.
%
%\section{Examples}
%We now give several examples of more sophisticated language constructs and how they may be implemented using this approach. We assume a type-level specification language structured like $\lamA$ but enriched with general inductive data kinds, though not necessarily dependent kinds. We assume kind inference when it is known to be decidable in System F (e.g. index kind declarations are omitted.) We also assume standard structures common to typed intermediate languages are available in the internal language. As in Section 2, missing else clauses are assumed to return \texttt{err}.
%
%\subsection{Record and Object Types}
%A record type can be characterized by a collection of static field names mapped to other types. They are commonly implemented using some flat, heterogenous, untagged data structure, because the offsets can be computed statically. We can therefore declare the type family of records as a type-level function that produces a type indexed by this specification and represented by a generic flat heterogeneous structure.
%
%\begin{verbatim}
%fun (spec : list[(str, Type)]) := 
%  type[RECORD spec] :: flat[map spec 
%    fun (s:str, t:Type) => (rep t)]
%\end{verbatim}
%
%To extract a value from a record, we need an appropriate operator. We can implement this as a type-level function that produces an operator when given an index string. We assume an isRec function has been defined to check that the representation type is consistent with the index as defined above. We also assume that type-level numbers can be translated to internal language numbers using the \verb|#| operator. Several other helper functions are also assumed where the meaning is apparent.
%
%\begin{verbatim}
%fun get(idx : str) := op(fun d : Den.
%  didxcase d of RECORD =>
%  if isRec typeof d =>
%  def spec = open idx typeof d
%  if containsIdx spec idx =>  
%    [[valof(d)[#(getOffset spec idx)], 
%      typeForIdx spec idx]])
%\end{verbatim}
%
%An analagous set function can be implemented. Object types are like records, except they have additional metadata expressing inheritance relationships and other annotations. This additional information could also be packaged into the type index in a straightforward way, allowing for languages supporting a variety of different object protocols.
%
%\subsection{Scientific Units}
%An example of a more specialized application domain is scientific unit checking. Numerical values (and sometimes collections of numerical values) can be annotated with dimensions (e.g. mass) and specific units of measure (e.g. grams). Addition and subtraction must operate on values with identical dimension and respect conversion rules if operating on heterogeneous units (e.g. grams + kilograms). Multiplication and division, on the other hand, cause the formation and elimination of compound units (e.g. meters/sec.). Although standalone type systems have been developed which respect these rules (e.g. \cite{conf/cefp/Kennedy09}), they are quite difficult to implement in terms of conventional language features, particularly if users hope to erase these annotations at run-time for performance reasons. Few languages have enough scientific users to justify adding this feature to the core language however.
%
%We can represent compound scientific types as being indexed by type-level lists of multiplicand and divisor units, where units are type-level values associated with dimensions. The corresponding rules can then be encoded directly as operators.
%
%\begin{verbatim}
%def unitNum = fun (mult div : list[SIUnit]) :=
%  type[SINUM mult, div] :: int
%\end{verbatim}
%\noindent
%We omit the details of the operator implementations.
%
%\subsection{Length-Indexed Lists}
%Length-indexed lists are a common example for dependently-typed languages. With type-level numbers, such lists are relatively straightforward to encode. All operators use standard arithmetic operations on the index to ensure that the length is accurately maintained.
%
%\begin{verbatim}
%def list = fun (length : nat, t : Type) := 
%  type[LIST length, t] :: array[rep t]
%\end{verbatim}
%
%Other common examples of dependently-typed data structures can be encoded similarly, though the dependency in this case is only on type-level values (or perhaps denotations), but not run-time values.
%
%\subsubsection{Interoperability Layers}
%A type-safe interoperability layer can be derived from an unsafe interoperability layer encoded by implementing the type system of the target language, with type representations corresponding to the core data structures available in the unsafe interoperability layer.
%\section{Related Work}
%\subsection{Type-Level Computation} %Haskell, Ur and $\Omega$mega
%System $\text{F}_{\omega}$ with simple case analysis provides the basis of type-level computation in Haskell (where type-level functions are called type families \cite{Chakravarty:2005:ATC}). Ur uses type-level records and names to support typesafe metaprogramming, with applications to web programming \cite{conf/pldi/Chlipala10}. $\Omega$mega adds algebraic data types at the type-level, using these to increase the expressive power of algebraic data types at the expression level \cite{conf/cefp/SheardL07}. Dependently-typed languages blur the traditional phase separation between types and expressions, so type-level computation is often implicitly used (though not always in its most general form, e.g. Deputy \cite{conf/icfp/ChenX05}, ATS \cite{conf/esop/ConditHAGN07}.)
%
%\subsection{Run-Time Indirection}
%{\it Operator overloading} \cite{vanWijngaarden:Mailloux:Peck:Koster:Sintzoff:Lindsey:Meertens:Fisker:acta:1975} and {\it metaobject dispatch} \cite{Kiczales91} are run-time protocols that translate operator invocations into function calls. The function is typically selected according to the type or value of one or more operands. These protocols share the notion of {\it inversion of control} with type-level specification. However, type-level specification is a {\it compile-time} protocol focused on enabling specialized verification and implementation strategies, rather than simply enabling run-time indirection.
%
%\subsection{Term Rewriting Systems}
%Many languages and tools allow developers to rewrite expressions according to custom rules. These can broadly be classified as {\it term rewriting systems}. Macro systems, such as those characteristic of the LISP family of languages \cite{mccarthy1978history}, are the most prominent example. Some compile-time metaprogramming systems also allow users to manipulate syntax trees (e.g. MetaML \cite{Sheard:1999:UMS}), and external rewrite systems also exist for many languages.
%These facilities differ from type-level specification in one or more of the following ways:
%
%\begin{enumerate}
%\item In type-level specification, the type of a value is determined separately from its representation; in fact, the same representation may be generated by multiple types. 
%\item We draw a distinction between the metalanguage, used to specify types and compile-time logic, the source grammar, used to describe run-time behavior, and the internal language, used to implement this behavior. Term rewriting systems generally do not draw this distinction. By doing so, each component language can be structured and constrained as appropriate for its distinct role, as we show.
%%\item With type-level specification, dispatch to a type-level function occurs implicitly on the basis of the structure of an expression. In contrast, most term-rewriting systems operate by  explicit invocation of a macro or specialized syntax. Some LISP macro systems have explored pattern-based dispatch (e.g. A*\cite{fowler2010domain}, EPP\cite{fowler2010domain}) and macro systems for object-oriented languages, like OpenC++ \cite{fowler2010domain} and OpenJava \cite{fowler2010domain}, do offer a somewhat limited form of operation-based dispatch.
%\item Many common macro systems and metaprogramming facilities operate at run-time. Compilers for some forms of LISP employ aggressive compile-time specialization techniques to attempt to minimize this overhead. Static and staged term-rewriting systems also exist (e.g. OpenJava\cite{TatM:OpenJCBMSJ}, Template Haskell\cite{SheardPeytonJones:Haskell-02}, MetaML \cite{Sheard:1999:UMS} and others). 
%\end{enumerate}
%
%\subsection{Language Frameworks}
%When the mechanisms available in an existing language prove insufficient, researchers and domain experts must design a new language. A number of tools have been developed to assist with this task, including compiler generators, language workbenches and domain-specific language frameworks (cf \cite{fowler2010domain}).
%
%A major barrier to adoption is the fact that interoperability is intrinsically problematic. Even languages which target a common platform, such as the Java Virtual Machine, can only interact using its limited set of primitives. Specialized typing rules are not checked at language boundaries, performance often suffers, and the syntax can be unnatural, particularly for languages which differ significantly from the platform's native language (e.g. Java).
%
%Instead of focusing on defining standalone languages, type-level specification gives greater responsibility in a granular manner to libraries. In this way, a range of constructs can coexist within the same program and, assuming that it can be shown by some method that various constructs are safely composable, be mixed and matched. The main limitation is that the protocol requires defining a fixed source grammar, whereas a specialized language has considerable flexibility in that regard. Nevertheless, as $\lamA$ shows, a simple grammar can be used quite flexibly.
%\subsection{Extensible Compilers}
%An alternative methodology is to implement language features granularly as compiler extensions. As discussed in Section 1, existing designs suffer from the same problems related to composability, modularity\-, safety and security as extensible languages, while also adding the issue of language fragmentation.
%
%Type-level specification can in fact be implemented within a compiler, rather than provided as a core language feature. This would resolve some of the issues, as described in this paper. However, by leveraging type-level computation to integrate the protocol directly into the language, we benefit from common module systems and other shared infrastructure. We also avoid the fragmentation issue.
%\subsection{Combining Logics and Specification Languages}
%Programming language specifications can be understood as formal logics. The question of {\it combining logics} \cite{sep-logic-combining} can be formulated as follows: given two logics, $\mathcal{L}_{1}$ and $\mathcal{L}_{2}$, how can we produce a minimal new logic, $\mathcal{L}$, such that it includes both $\mathcal{L}_{1}$ and $\mathcal{L}_{2}$?
%
%The major approaches are based in category theory and include {\it fibring} \cite{acs:css:ccal:98a} and the framework of {\it institutions} \cite{Diaconescu-Futatsugi01}. Many variants of these approaches exist, addressing different formulation styles and focusing on guaranteeing different properties about the combined logics by introducing various constraints.
%
%Several {\it specification languages} (or {\it logical frameworks}) based on these theoretical formulations exist, including the OBJ family of languages (e.g. CafeOBJ \cite{Diaconescu-Futatsugi01}). They provide support for verifying a program against a language specification, and can automatically execute these programs as well in some cases. The  language itself specifies which verification and execution strategies are used.
%
%Type-determined compilation takes a more concrete approach to the problem, focusing on combining {\it implementations} of different\- logics, rather than simply their specifications. In other words, it focuses on combining {\it type checkers} and {\it implementation strategies} rather than more abstract representations of a language's type system and dynamic semantics. In Section 4, we outlined a preliminary approach based on proof assistant available for the type-level language to unify these approaches, and we hope to continue this line of research in future work.
%
%
%\section{Future Work}
%\subsection{General-Purpose Language Development}
%We have started to explore applications of this work to scientific programming, building a variant of this system (with a weaker metalanguage than described in Section 4) that targets the OpenCL programming language for GPU computing. The entirety of the C99 can be implemented using a variant of this approach that is not strictly compositional (an approach that can lead to non-termination). We will continue this work while also continuing to develop the more theoretically sound foundations described here.
%
%Although we have implemented several powerful data structures, more thorough  case studies must be performed with such a language before we can definitively say that this approach could lead to a practical programming language. We conjecture that this process will motivate further research questions related to this protocol as well.
%
%\subsection{Integration with Tools}
%Developers increasingly turn to integrated development environments, debuggers and refactoring tools to assist with their programming tasks. Many of these tools are built against a fixed language definition. There remain significant challenges in understanding how to integrate language designs based on type-level specification into these tools effectively.
%
%\section{Conclusion}
%In conclusion, we have shown a new new approach to language extensibility that works at the type-level. This approach is flexible but can also shown to be safe, in that it respects global safety properties of the language and does not negative impact the integrity of compilation. We gave a simple core calculus for this approach that we hope will be useful for distilling the essential concept, sketched an extension of the calculus that can be used to prove stronger metatheoretic guarantees and sketched several examples of relatively complex language constructs that admit implementation using this approach.
%
%\acks
%%The author is grateful to Jonathan Aldrich, Robert Bocchino and anonymous referees for their useful suggestions. This work was funded by the DOE Computational Science Graduate Fellowship under grant number DE-FG02-97ER25308.
%%Acknowledgments, if needed.
%
%% We recommend abbrvnat bibliography style.
%
%\bibliographystyle{abbrvnat}
%
%% The bibliography should be embedded for final submission.
%
%\bibliography{../research}
%%\softraggedright
%%P. Q. Smith, and X. Y. Jones. ...reference text...
%
%%\appendix
%%The grammar of $\lamTDC$ is:
%%\begin{eqnarray*}
%%\kappa & ::= & \kinds\\
%%\tau & ::= & \metaterms\\
%%e & ::= & \sourcegrammar\\
%%\tvarCtx & ::= & \datatypes\\
%%\evarCtx & ::= & \ilgrammar
%%\end{eqnarray*}
%%Note that $\intlit$ is a placeholder for any integer literal and $x$ is a placeholder for variable names. In this paper, we use strings of italicized, lowercase letters (e.g. {\it gram}).
%%\subsection{Intermediate Language Statics}
%%Intermediate language terms, $\evarCtx$, form a simple expression language. The intermediate language's types, $\tvarCtx$, are called {\it data types}. The statics take the form $\evarCtx :: \tvarCtx$.
%%\begin{mathpar}
%%\intlitT
%%~~~~
%%\addT
%%\end{mathpar}\begin{mathpar}
%%\tupleT\end{mathpar}\begin{mathpar}
%%\fstT
%%~~~~
%%\sndT
%%~~~~
%%\faultT
%%\end{mathpar}
%%\subsection{Metalanguage Statics}
%%Type-level terms, $\tau$, form a simply typed lambda calculus. The types of the metalanguage, $\kappa$, are called {\it kinds}. 
%%Statics take the form $\entails{\ctx}{\tau \sim \kappa}$, where the context, $\ctx$, operates as a set of assignments of kinds to variables and $\ctx \uplus x \sim \tau$ replaces any previous assignment for $x$ in $\ctx$ with $x\sim\tau$.
%%\paragraph{Standard Terms}
%%\newcommand{\axiom}[1]{\inferrule{ }{#1}}
%%\begin{mathpar}
%%\varT~~~~
%%\kfnIT\\
%%\kfnET
%%\end{mathpar}
%%For convenience, a simple ${\sf let}$ form is included.
%%\begin{mathpar}
%%\letT
%%\end{mathpar}
%%The ${\sf if}$ form checks whether two terms are $\beta$-equivalent. The two branches must evaluate to terms of the same kind.
%%\begin{mathpar}
%%\ifT
%%\end{mathpar}
%%The remaining statics are described in Section 3. We list them again, and give each a name, below.
%%\paragraph{Type Construction}
%%\begin{mathpar}
%%\typeT\\
%%\errT
%%\end{mathpar}
%%\paragraph{Code Generation Primitives}
%%\begin{mathpar}
%%\ilrepT\\
%%\iladdT\\
%%\iltupleT\\
%%\ilfstT~~~~
%%\ilsndT
%%\end{mathpar}
%%
%%\paragraph{Source Parsing Rules}
%%\begin{mathpar}
%%\numtreePT~~~~\newtreePT\\
%%\optreePT
%%\end{mathpar}
%%\paragraph{Syntax Tree Representations}
%%\begin{mathpar}
%%\numtreeT\\\newtreeT\end{mathpar}{\begin{mathpar}\optreeT\\\visitT
%%\end{mathpar}
%%\paragraph{Source Program Representation}
%%\begin{mathpar}
%%\sourcetT
%%\end{mathpar}
%%\subsection{Metalanguage Dynamics}
%%We now list the dynamics of the metalanguage.
%%\paragraph{Standard Terms}
%%\begin{mathpar}
%%\inferrule{ }{
%%	\final{\kfn{x}{\kappa}{\tau}}
%%}
%%~~~~
%%\inferrule{
%%	\tau_0 \reduces \tau_0'
%%}{
%%	\tau_0~\tau_1 \reduces \tau_0'~\tau_1
%%}
%%\\
%%\inferrule{ }{
%%	(\kfn{x}{\kappa}{\tau})~\tau_0 \reduces \subst{\tau_0}{x}{\tau}
%%}
%%~~~~
%%\inferrule{ }{
%%	\klet{x}{\tau_0}{\tau_1} \reduces \subst{\tau_0}{x}{\tau_1}
%%}\\
%%\inferrule{\tau_0 \reduces \tau_0'}{
%%	\ifeq{\tau_0=\tau_1}{\tau_\true}{\tau_\false} \reduces \ifeq{\tau_0' = \tau_1}{\tau_\true}{\tau_\false}
%%}\\
%%\inferrule{\final{\tau_0}\\\tau_1 \reduces \tau_1'}{
%%	\ifeq{\tau_0=\tau_1}{\tau_\true}{\tau_\false} \reduces \ifeq{\tau_0 = \tau_1'}{\tau_\true}{\tau_\false}
%%}\\
%%\inferrule{\final{\tau_0}\\\final{\tau_1}\\\tau_0 = \tau_1}{
%%	\ifeq{\tau_0=\tau_1}{\tau_\true}{\tau_\false} \reduces \tau_\true
%%}\\
%%\inferrule{\final{\tau_0}\\\final{\tau_1}\\\tau_0 \neq \tau_1}{
%%	\ifeq{\tau_0=\tau_1}{\tau_\true}{\tau_\false} \reduces \tau_\false
%%}
%%\end{mathpar}
%%$\tau_0 = \tau_1$ and $\tau_0 \neq \tau_1$ denote syntactic equality and inequality.
%%\paragraph{Type Construction}
%%\begin{mathpar}
%%\inferrule{ }{
%%	\final{\typedefstd}
%%}
%%~~~~
%%\inferrule{ }{
%%	\final{\err}
%%}
%%\end{mathpar}
%%\paragraph{Code Generation Primitives}
%%\begin{mathpar}
%%\inferrule{ }{
%%	\final{\evarCtx[\evarCtx]}
%%}
%%~~~~
%%\inferrule{
%%	\tau_L \reduces \tau_L'
%%}{
%%	\evarCtx_{+}(\tau_L, \tau_R) \reduces \evarCtx_{+}(\tau_L', \tau_R)
%%}\end{mathpar}\begin{mathpar}
%%\inferrule{
%%	\tau_R \reduces \tau_R'	
%%}{
%%	\evarCtx_{+}(\evarCtx[\evarCtx_L], \tau_R) \reduces \evarCtx_{+}(\evarCtx[\evarCtx_L], \tau_R')
%%}\end{mathpar}\begin{mathpar}
%%\inferrule{ }{
%%	\evarCtx_{+}(\evarCtx[\evarCtx_L], \evarCtx[\evarCtx_R]) \reduces \evarCtx[\evarCtx_L + \evarCtx_R]
%%}\end{mathpar}\begin{mathpar}
%%\inferrule{
%%	\tau_L \reduces \tau_L'
%%}{
%%	\evarCtx_{\times}(\tau_L, \tau_R) \reduces \evarCtx_{\times}(\tau_L', \tau_R)
%%}\end{mathpar}\begin{mathpar}
%%\inferrule{
%%	\tau_R \reduces \tau_R'	
%%}{
%%	\evarCtx_{\times}(\evarCtx[\evarCtx_L], \tau_R) \reduces \evarCtx_{\times}(\evarCtx[\evarCtx_L], \tau_R')
%%}\end{mathpar}\begin{mathpar}
%%\inferrule{ }{
%%	\evarCtx_{\times}(\evarCtx[\evarCtx_L], \evarCtx[\evarCtx_R]) \reduces \evarCtx[(\evarCtx_L, \evarCtx_R)]
%%}\end{mathpar}\begin{mathpar}
%%\inferrule{\tau_\times \reduces \tau_\times'}{
%%	\evarCtxfst{\tau_\times} \reduces \evarCtxfst{\tau_\times'}
%%}
%%~~~~
%%\inferrule{ }{
%%	\evarCtxfst{\evarCtx[(\evarCtx_L, \evarCtx_R)]} \reduces \evarCtx[\evarCtx_L]
%%}\end{mathpar}\begin{mathpar}
%%\inferrule{\tau_\times \reduces \tau_\times'}{
%%	\evarCtxsnd{\tau_\times} \reduces \evarCtxsnd{\tau_\times'}
%%}
%%~~~~
%%\inferrule{ }{
%%	\evarCtxsnd{\evarCtx[(\evarCtx_L, \evarCtx_R)]} \reduces \evarCtx[\evarCtx_R]
%%}
%%\end{mathpar}
%%\paragraph{Syntax Tree Representations}
%%\begin{mathpar}
%%\inferrule{
%%	\tau_n \reduces \tau_n'
%%}{
%%	\numtree{\tau_n} \reduces \numtree{\tau_n'}
%%}~~~~
%%\axiom{\final{\numtree{\evarCtx[\intlit]}}}\end{mathpar}\begin{mathpar}
%%\inferrule{
%%	\tau_\star \reduces \tau_\star'
%%}{
%%	\newtreestd \reduces \newtree{\tau_\star'}{\tau_\pi}
%%}~~~~
%%\inferrule{
%%	\final{\tau_\star}\\
%%	\tau_\pi \reduces \tau_\pi'
%%}{
%%	\newtreestd \reduces \newtree{\tau_\star}{\tau_\pi'}
%%}\end{mathpar}\begin{mathpar}
%%\inferrule{
%%	\final{\tau_\star}\\
%%	\final{\tau_\pi}
%%}{
%%	\final{\newtreestd}
%%}\end{mathpar}\begin{mathpar}
%%\inferrule{
%%	\tauL \reduces \tauL'
%%}{
%%	\optreestd \reduces \optree{\tauL'}{\tauR}
%%}~~~~
%%\inferrule{
%%	\final{\tauL}\\
%%	\tauR \reduces \tauR'
%%}{
%%	\optreestd \reduces \optree{\tauL}{\tauR'}
%%}\end{mathpar}\begin{mathpar}
%%\inferrule{
%%	\final{\tauL}\\
%%	\final{\tauR}
%%}{
%%	\final{\optreestd}
%%}
%%\end{mathpar}
%%\begin{mathpar}
%%\inferrule{
%%	\taupi \reduces \taupi'
%%}{
%%	\kvisitstd \reduces \kvisit{\taupi'}{\tau_{\#}}{\tauadd}{\taunew}
%%}\end{mathpar}\begin{mathpar}
%%\axiom{\kvisit{\numtreestd}{\tau_{\#}}{\tauadd}{\taunew} \reduces \tau_{\#}~\tau_n}\end{mathpar}\begin{mathpar}
%%\axiom{\kvisit{\newtreestd}{\tau_{\#}}{\tauadd}{\taunew} \reduces \taunew~\tau_\star~\taupi}\end{mathpar}\begin{mathpar}
%%\axiom{\kvisit{\optreestd}{\tau_{\#}}{\tauadd}{\taunew} \reduces \tauadd~\tauL~\tauR}\end{mathpar}
%%\paragraph{Source Program Representations}
%%\begin{mathpar}
%%\axiom{\final{\source{\intlit}}}\\
%%\inferrule{
%%	\tau \reduces \tau'
%%}{
%%	\source{\newobj{\tau}{e}} \reduces \source{\newobj{\tau'}{e}}
%%}\\
%%\inferrule{
%%	\final{\tau}\\
%%	\source{e} \reduces \source{e'}
%%}{
%%	\source{\newobj{\tau}{e}} \reduces \source{\newobj{\tau}{e'}}
%%}\\
%%\inferrule{
%%	\final{\tau}\\
%%	\final{\source{e}}
%%}{
%%	\final{\source{\newobj{\tau}{e}}}
%%}\\
%%\inferrule{
%%	\source{e_L} \reduces \source{e_L'}
%%}{
%%	\source{\absop{e_L}{e_R}} \reduces \source{\absop{e_L'}{e_R}}
%%}\\
%%\inferrule{
%%	\final{\source{e_L}}\\
%%	\source{e_R} \reduces \source{e_R'}
%%}{
%%	\source{\absop{e_L}{e_R}} \reduces \source{\absop{e_L}{e_R'}}
%%}\\
%%\inferrule{
%%	\final{\source{e_L}}\\
%%	\final{\source{e_R}}
%%}{
%%	\final{\source{\absop{e_L}{e_R}}}
%%}
%%\end{mathpar}
%%\paragraph{Substitution Dynamics}
%%\subparagraph{Standard Terms}
%%\begin{mathpar}
%%\inferrule{ }{
%%	\subst{\tau}{x}{x} \reduces \tau
%%}
%%~~~~
%%\inferrule{x \neq y}{
%%	\subst{\tau}{x}{y} \reduces y
%%}\end{mathpar}\begin{mathpar}
%%\inferrule{ }{
%%	\subst{\tau}{x}{\kfn{x}{\kappa}{\tau_0}} \reduces \kfn{x}{\kappa}{\tau_0}
%%}\end{mathpar}\begin{mathpar}
%%\inferrule{
%%	x \neq y\\
%%	\subst{\tau}{x}{\tau_0} \reduces \tau_0'	
%%}{
%%	\subst{\tau}{x}{\kfn{y}{\kappa}{\tau_0}} \reduces \kfn{y}{\kappa}{\tau_0'}
%%}\end{mathpar}\begin{mathpar}
%%\inferrule{
%%	\subst{\tau}{x}{\tau_0} \reduces \tau_0'\\
%%	\subst{\tau}{x}{\tau_1} \reduces \tau_1'
%%}{
%%	\subst{\tau}{x}{\tau_0~\tau_1} \reduces \tau_0'~\tau_1'
%%}\end{mathpar}\begin{mathpar}
%%\inferrule{
%%	\subst{\tau}{x}{\tau_0} \reduces \tau_0'\\
%%}{
%%	\subst{\tau}{x}{\klet{x}{\tau_0}{\tau_1}} \reduces \klet{x}{\tau_0'}{\tau_1}
%%}\end{mathpar}\begin{mathpar}
%%\inferrule{
%%	x \neq y\\
%%	\subst{\tau}{x}{\tau_0} \reduces \tau_0'\\
%%	\subst{\tau}{x}{\tau_1} \reduces \tau_1'
%%}{
%%	\subst{\tau}{x}{\klet{y}{\tau_0}{\tau_1}} \reduces \klet{y}{\tau_0'}{\tau_1'}
%%}\end{mathpar}\begin{mathpar}
%%\inferrule{
%%	\subst{\tau}{x}{\tau_0} \reduces \tau_0'~~~~
%%	\subst{\tau}{x}{\tau_1} \reduces \tau_1'~~~~
%%	\subst{\tau}{x}{\tau_\true} \reduces \tau_\true'~~~~
%%	\subst{\tau}{x}{\tau_\false} \reduces \tau_\false'
%%}{
%%	\subst{\tau}{x}{\ifeq{\tau_0 = \tau_1}{\tau_\true}{\tau_\false}} \reduces \ifeq{\tau_0' = \tau_1'}{\tau_\true'}{\tau_\false'}
%%}\end{mathpar}
%%\subparagraph{Type Construction}
%%\begin{mathpar}
%%\inferrule{
%%	\subst{\tau}{x}{\taunewg} \reduces \hat{\tau}_{\sf new}^\evarCtx\\
%%	\subst{\tau}{x}{\tauaddg} \reduces \hat{\tau}_\diamond^\star\\
%%	\subst{\tau}{x}{\tauaddt} \reduces \hat{\tau}_\diamond^\evarCtx
%%}{
%%	\subst{\tau}{x}{\typedefstd} \reduces \subst{\tau}{x}{\typedef{a}{\hat{\tau}_{\sf new}^\evarCtx}{\hat{\tau}_\diamond^\star}{\hat{\tau}_\diamond^\evarCtx}{\tvarCtxmap}{f}{g}}
%%}\end{mathpar}\begin{mathpar}
%%\inferrule{ }{
%%	\subst{\tau}{x}{\err} \reduces \err
%%}\end{mathpar}
%%\subparagraph{Code Generation Primitives}
%%\begin{mathpar}
%%\inferrule{ }{
%%	\subst{\tau}{x}{\evarCtx[\evarCtx]} \reduces \evarCtx[\evarCtx]
%%}\end{mathpar}\begin{mathpar}
%%\inferrule{
%%	\subst{\tau}{x}{\tau_L} \reduces \tau_L'\\
%%	\subst{\tau}{x}{\tau_R} \reduces \tau_R'
%%}{
%%	\subst{\tau}{x}{\evarCtx_{+}(\tau_L, \tau_R)} \reduces \evarCtx_{+}(\tau_L', \tau_R')
%%}\end{mathpar}\begin{mathpar}
%%\inferrule{
%%	\subst{\tau}{x}{\tau_L} \reduces \tau_L'\\
%%	\subst{\tau}{x}{\tau_R} \reduces \tau_R'
%%}{
%%	\subst{\tau}{x}{\evarCtx_{\times}(\tau_L, \tau_R)} \reduces \evarCtx_{\times}(\tau_L', \tau_R')
%%}\end{mathpar}\begin{mathpar}
%%\inferrule{
%%	\subst{\tau}{x}{\tau_\times} \reduces \tau_\times'
%%}{
%%	\subst{\tau}{x}{\evarCtxfst{\tau_\times}} \reduces \evarCtxfst{\tau_\times'}
%%}
%%~~~~
%%\inferrule{
%%	\subst{\tau}{x}{\tau_\times} \reduces \tau_\times'
%%}{
%%	\subst{\tau}{x}{\evarCtxsnd{\tau_\times}} \reduces \evarCtxsnd{\tau_\times'}
%%}\end{mathpar}
%%\subparagraph{Syntax Tree Representations}
%%\begin{mathpar}
%%\inferrule{
%%	\subst{\tau}{x}{\tau_n} \reduces \tau_n'
%%}{
%%	\subst{\tau}{x}{\numtreestd} \reduces \numtree{\tau_n'}
%%}
%%~~~~
%%\inferrule{
%%	\subst{\tau}{x}{\tau_\star} \reduces \tau_\star'\\
%%	\subst{\tau}{x}{\tau_\pi} \reduces \tau_\pi'
%%}{
%%	\subst{\tau}{x}{\newtreestd} \reduces \newtree{\tau_\star'}{\tau_\pi'}
%%}\end{mathpar}\begin{mathpar}
%%\inferrule{
%%	\subst{\tau}{x}{\tau_L} \reduces \tau_L'\\
%%	\subst{\tau}{x}{\tau_R} \reduces \tau_R'
%%}{
%%	\subst{\tau}{x}{\optreestd} \reduces \optree{\tau_L'}{\tau_R'}
%%}\end{mathpar}\begin{mathpar}
%%\inferrule{
%%	\subst{\tau}{x}{\tau_\pi} \reduces \tau_\pi'~~~~
%%	\subst{\tau}{x}{\tau_{\#}} \reduces \tau_{\#}'\\
%%	\subst{\tau}{x}{\tau_{\sf new}} \reduces \tau_{\sf new}'~~~~
%%	\subst{\tau}{x}{\tau_\diamond} \reduces \tau_\diamond
%%}{
%%	\subst{\tau}{x}{\kvisit{\tau_\pi}{\tau_{\#}}{\tau_\diamond}{\tau_{\sf new}}} \reduces
%%	\kvisit{\tau_\pi'}{\tau_{\#}'}{\tau_\diamond'}{\tau_{\sf new}'}
%%}\end{mathpar}
%%\subparagraph{Source Program Representations}
%%\begin{mathpar}
%%\inferrule{ }{
%%	\subst{\tau}{x}{\source{\intlit}} \reduces \source{\intlit}
%%}\end{mathpar}\begin{mathpar}
%%\inferrule{
%%	\subst{\tau}{x}{\tau_\star} \reduces \tau_\star'\\
%%	\subst{\tau}{x}{\source{e}} \reduces \source{e'}
%%}{
%%	\subst{\tau}{x}{\source{\newobj{\tau_\star}{e}}} \reduces \source{\newobj{\tau_\star'}{e'}}
%%}\end{mathpar}\begin{mathpar}
%%\inferrule{
%%	\subst{\tau}{x}{\source{e_L}} \reduces \source{e_L'}\\
%%	\subst{\tau}{x}{\source{e_R}} \reduces \source{e_R'}
%%}{
%%	\subst{\tau}{x}{\source{\absop{e_L}{e_R}}} \reduces \source{\absop{e_L'}{e_R'}}
%%}
%%\end{mathpar}
%%\subsection{Canonical Forms}
%%\begin{center}
%%\begin{tabular}{ | c | c | }
%%  \hline            
%%  \textbf{Kind} & \textbf{Canonical Form(s)} \\
%%  \hline
%%  $\kappa_1 \rightarrow \kappa_2$ & $\kfn{x}{\kappa}{\tau}$ \\
%%  $\star$ & $\typedefstd$, $\err$ \\
%%  $\sfype{\intt}$ & $\evarCtx[\intlit]$ \\
%%  $\sfype{\tvarCtx_{L} \times \tvarCtx_{R}}$ & $\evarCtx[(\evarCtx_{L}, \evarCtx_{R})]$ where $\evarCtx[\evarCtx_{L}]$ is a canonical form for $\sfype{\tvarCtx_{L}}$ \\ & and $\evarCtx[\evarCtx_{R}]$ is a canonical form for $\sfype{\tvarCtx_{R}}$.\\
%%  $\sfype{\otimes}$ & $\evarCtx[\fault]$\\
%%  $\parsetree$ & $\numtreestd$, $\newtreestd$, $\optreestd$\\
%%  $\sourcet$ & $\source{e}$\\
%%  \hline  
%%\end{tabular}
%%\end{center}
\end{document}

### 2013-04-11 globaldsl13/approach.tex
Added GlobalDSL draft

\section{Approach}
\label{s:approach}

Figure~\ref{f:dsl-if} illustrates our approach to type-based extension for the example in Figure 1.  The class \verb|ComponentArch| defines a grammar extension to the language's expression grammar. This grammar must not conflict with the fixed base expression grammar, but because it is only used when an expression of the type \verb|ComponentArch| type is requested \emph{and} a whitespace-delimited block has been introduced, there is no need to guarantee that this grammar does not interfere with other type-associated grammars.

The declared grammar uses a standard BNF notation except for the declaration \verb|EXP of URL| on Line 7. This indicates that at this position in the grammar, a term of type \verb|URL| is needed. Thus, the domain-specific grammar (if any) associated with the \verb|URL| class may be used here. Because URLs are typically short, it may be undesirable to require a new whitespace-delimited block, so the grammar defines blocks delimited by brackets as equivalent to whitespace-delimited blocks as well, as can be seen being used in Figure 1.

\begin{figure}
  \centering\small
  \begin{lstlisting}[language=Python,morekeywords={grammar,where,ID,TYPE,EXP}]
class ComponentArch
  grammar ::= ((component|policy) \n)+ where 
    component ::= 
      ("external ")? "component " ID ":" TYPE
        [componentProperty*]?
    componentProperty ::= 
        "location "(EXP of URL)
      | "connects to" (ID ",")* ID
    ...
  \end{lstlisting}
  \caption{Type-Associated Grammar Definition}
  \label{f:dsl-if}
\end{figure}

%In this example, we first parse the if condition using a standard
%expression parser \texttt{StdExpParser}.  The expression parser is
%passed a sub-stream that contains only tokens preceding the end of
%the line.  After parsing the condition expression, we then accept an
%\texttt{INDENT} token, indicating that we are now expecting one or
%more indented lines.  This token is consumed by using a DSL for
%parsing, which is modeled after BNF syntax.  The definition of this
%parsing DSL is not shown, but it can be defined using the same approach.
%%described in this section.
%
%After that, we use a standard statement parser to parse the indented lines as
%a series of one or more statements and construct and return
%an \texttt{IfAST} object representing the if statement.  The
%\texttt{IfAST} class is an ordinary class that implements the
%\texttt{AST} interface, which requires the class to define its semantics via
%an interpreter or via translation to lower-level constructs.  The
%details of the AST interface are out of scope here.  Once the
%semantics of \keyw{if} are defined, we can use it as shown in
%the \texttt{abs} method, which computes the absolute value of its
%argument.
%
%
\paragraph{Parsing and Typechecking}
Wyvern source code is parsed using a standard whitespace-delimited lexer followed by an initial typechecking and parsing phase that occurs in tandem.  The lexer for Wyvern generates a stream of tokens structured hierarchically, where indentation (and the equivalent bracket notation) increases the depth of the stream. Indentation and de-indentation can thus be simply  understood as a special pair of tokens, \texttt{INDENT} and \texttt{DEDENT}, generated by the lexer and delimiting portions of the stream that will be handled in a special way during the next phase.

The standard Wyvern parser reads at a single level of the stream at a time. Initially, it parses the stream as a term of sort \verb|Exp|, which represents the core expressions and declaration forms of the language. This sort includes basic facilities for declaring and working with functions and variables, objects and perhaps other to-be-determined core data types. When an indented block is encountered, however, it is left in-place within the parsed syntax tree.

When the parsing phase at a particular level is complete, it is typechecked according to a largely conventional protocol. In Wyvern, the type of a function is determined by its declaration, and cannot be affected by its use (this constraint may be relaxed in some instances in future iterations of the language). Thus, when a function is being applied, the expected types of its arguments are known. If, at a particular location where a function argument was expected, an indented block is found, then it is parsed not according to the \verb|Exp| grammar, but according to the \emph{type-associated grammar} contained in that type's definition. Figure 4 shows an example of such a grammar. That grammar may contain productions of the form $\texttt{EXP of }\tau$ where $\tau$ is another type in scope. A delimited block must be present at this location as well, and it will be parsed by the type-associated grammar of $\tau$.

A particular function application can contain only one \emph{primary} indented block. If multiple arguments support DSL syntax, then all those other than the first in the argument ordering must be invoked by a partially-indented keyword. This can be seen in Figure \ref{f:dsl-multi}. Note that this partially-indented keyword ends the first delimited block and begins a new block for the keyword argument \verb|toServer| of the function \verb|db.sendQuery|. This can be seen as similar in form to keywords in functions in Smalltalk \cite{smalltalk}.

\paragraph{Keyword-Directed Invocation}
Keyword-directed invocation of a DSL is simply a special case of this approach. In particular, a keyword can be defined as a function with a single argument of a type specific to that keyword. The type contains the implementation of the domain-specific syntax associated with that keyword. If the construction of this type is controlled such that it can only be via the domain-specific syntax (and not any equivalent syntax in the core language), then it is precisely equivalent to keyword-directed invocation of a DSL.

As keyword-directed DSL invocation is highly explicit, we allow keywords to appear not just at the end of a line, but also within expressions.  In this case, the keyword determines how much of the expression (and potentially following indented lines) that it will parse.

%Our approach supports not just key{\-}word-based extension, as in
%\keyw{if}, but also type-based syntax extensions.  Consider the
%database query DSL used in Figure~\ref{f:dsl-query}.  When parsing
%this code, the standard Wyvern parser reads the identifier
%\texttt{productDB} and discovers, based on the architecture from
%Figure~\ref{f:dsl-arch}, that \texttt{productDB} has type
%\texttt{Database}.  To support type-based syntax extension, the Wyvern
%parser examines this type, defined in Figure~\ref{f:dsl-define}, and
%discovers that the \texttt{Database} class object implements the
%\texttt{Parser} interface.  Note that we associate the parser with the
%\texttt{Database} \textit{class}, not the \texttt{Database}
%instance. This works well because the class is available at compile
%time, during parsing,
%but the instance is not available until execution.

\begin{figure}
\centering
\begin{lstlisting}
db.sendQuery
    select x where x > 10
  toServer
    www.server.org
\end{lstlisting}
\caption{Function applications involving multiple white-space delimited arguments use partially-indented keywords.}
\label{f:dsl-multi}
\end{figure}
%\begin{figure}
%  \centering
%  \begin{lstlisting}
%class Database class implements Parser
%	class meth parse(db : AST,
%	                 stream : TokStream):AST
%		if (stream ::= "where")
%			// parse where clause
%		else ... // parse other constructs
%	... // other definitions for Database
%
%// equivalent code for if using types
%class IfKeyType class implements Parser
%	class meth parse(ifKeyword : AST,
%	                 stream:ParseStream):AST
%		val cond = ... // same code as previous Figure
%	class prop instance = new IfKeyType // singleton
%val (@if@) = IfKeyType.instance
%  \end{lstlisting}
%  \caption{Type-based DSL definition}
%  \label{f:dsl-define}
%\end{figure}
%
\paragraph{Procedural Parsing}
The grammar definition in Figure 4 is declarative and relies on a simple parser generation framework included in Wyvern. It may be desirable for more sophisticated parsing algorithms to be encoded for particular domain-specific languages. For example, an interoperability layer with another full-scale programming language may wish to support parsing features not expressible in Wyvern's declarative grammar language. It can be observed, however, that a declarative grammar inside a class definition can be thought of as inducing a class method (a concept also borrowed from Smalltalk \cite{smalltalk}) called \verb|parse|, that transforms a stream of tokens to an AST representation. This interface can be exposed directly to programmers who wish to extend the parsing capabilities of the language in more sophisticated ways. We leave the details of the AST representation and its processing during later stages of compilation as future work.

### 2013-04-11 globaldsl13/conclusion.tex
Added GlobalDSL draft

\section{Conclusion} % and Future Work}
\label{s:conclusion}

In this paper, we described how extensible parsing in Wyvern makes for
a solid platform to support whitespace-delimited, type-directed embedded DSLs. In the
future, we aim to implement a wide variety of DSLs in Wyvern tweaking
our approach and implementation thereof to provide a comprehensive example of
supporting multiple interacting DSLs in a safe and easy-to-use manner.

% \todo{tie features to goals}

% \todo{implementation and validation plans}

### 2013-04-11 globaldsl13/globaldsl13.tex
Added GlobalDSL draft

\documentclass[english]{sigplanconf}
\usepackage{palatino}
\usepackage[T1]{fontenc}
\usepackage[latin9]{inputenc}
\usepackage{listings}
\usepackage{url}
\usepackage{babel}
\usepackage{graphicx}
%\usepackage{microtype}

%\lstset{
%  numbers=left,
%  basicstyle=\ttfamily,
%  tabsize=4,
%  frame=single,
%}

\lstset{tabsize=3, basicstyle=\ttfamily\small, commentstyle=\itshape\rmfamily, numbers=left, numberstyle=\tiny, language=java,moredelim=[il][\sffamily]{?},mathescape=true,showspaces=false,showstringspaces=false,columns=fullflexible,xleftmargin=15pt,escapeinside={(@}{@)}, morekeywords=[1]{application,policy,external,component,connects,to,meth,let,in,val,where,return,group,by,within,connect,with,attr,html,head,title,style,body,div,keyword,location,schema,must,salt,select,text}}
\lstloadlanguages{Java,VBScript,XML,HTML,Python}

\newcommand{\todo}[1]{\textbf{[TODO: #1]}}
\newcommand{\keyw}[1]{\texttt{\textbf{#1}}}

\begin{document}

\title{Whitespace-Delimited, Type-Directed Parsing for Embedded DSLs}

\authorinfo{Cyrus Omar, Benjamin Chung$^{1}$, Darya Kurilova, Alex Potanin$^{2}$,
  and Jonathan Aldrich}
{Carnegie Mellon University}
{\{comar, darya, aldrich\}@cs.cmu.edu, bwchung@andrew.cmu.edu$^{1}$, and alex@ecs.vuw.ac.nz$^{2}$}
% Maybe just keep my cs account active after I leave for a few years?
% Otherwise: alex@ecs.vuw.ac.nz is my email...
\maketitle

\begin{abstract}

Domain-specific languages enhance ease of use, expressiveness and
verifiability, 
but defining and using different 
DSLs within a single application remains difficult.  
We present an approach to embedded DSLs where:
1) a whitespace-delimited lexer is shared across DSLs, and 
2) the parsing and type checking phases occur in tandem so that types can define domain-specific parsing strategies governing whitespace-delimited blocks.
We argue that this approach occupies a sweet spot with both 
high expressiveness and ease of use. We outline a design
and ongoing implementation of this strategy
and examine the relationship between the type-directed and keyword-directed strategies for parsing of embedded DSLs.

\end{abstract}

\begin{sloppypar}

\input{intro}
\input{motivation}
\input{approach}
\input{implementation}
\input{related}
\input{conclusion}

%TODO TODO TODO TODO TODO TODO TODO TODO add acknowledgments here!!!!!!  ARO

\end{sloppypar}

\bibliographystyle{plain}
\bibliography{biblio}
\end{document}

### 2013-04-11 globaldsl13/intro.tex
Added GlobalDSL draft

\section{Introduction}
\label{s:intro}

Domain-specific languages (DSLs) \cite{shellchapter, coldfusion,
  sinatra}
% have been widely-studied because they 
enable the expression of  
domain-specific ideas naturally and permit domain-specific
verification and compilation strategies.  For DSLs to reach their full
potential, however, it must be possible to easily use multiple
DSLs at once within a host general-purpose language (GPL), such that DSL code
and GPL glue code can inter-operate to form a complete application. More specifically, we propose the following essential design criteria:

\begin{itemize}

\item \emph {Composability}: It should be possible to use multiple DSLs and a GPL
%, in addition to a GPL,
within a single program unit.  %Here 
Within the text file-based paradigm familiar to most developers, this 
means including multiple DSLs within a single file.
 Moreover, it should be possible to embed code written in one DSL
  within another DSL without them requiring knowledge of each other or interfering with one another. So, more specifically, DSLs should be \emph{safely composable}.

\item \emph{Interoperability}: DSLs should be able to \emph{interoperate}. A necessary condition for interoperability is that it should be
  possible to pass and operate on values 
%such as functions or data structures
  defined in foreign DSLs. Additional requirements, such as the ability to do so naturally (rather than requiring a large amount of ``glue code'') and safely may also be relevant in many settings. 
%Moreover,
  Minimally, however, if one DSL defines a function or data structure satisfying an interface specified in
  a common interface description language (such as the type system of
  a host GPL), code written in another DSL should be able to use those
  values according to that interface, without requiring that the client DSL
  have knowledge of the details of the provider DSL or \emph{vice versa}.

\end{itemize}

%\item type system: one DSL depends on types defined by another DSL,
%  can use objects of that type in special ways [this goes beyond the
%    scope - save for a later paper]

In addition to composability and interoperability, we believe that to be most useful, a system supporting DSLs should have the following characteristics:

\begin{itemize}

\item \emph{Flexibility}: Have flexible enough DSL syntax to support a wide variety of text-based notations;

\item \emph{Identifiability}: Make it easy for programmers to identify which code is written in which DSL;

\item \emph{Consistency}: Encourage DSLs written in similar styles in order to enhance readability and learnability of each DSL;

\item \emph{Simplicity}: Keep the cost of defining a DSL low.

%\item Share conventions between DSLs and a host language, making each DSL easier for programmers to learn, helping programmers to identify which code is written in which DSL, and avoiding unintended conflicts between DSLs. \todo{Avoid conflicts (visual and real), enhance learnability}

%\item Reuse low-level mechanisms and design decisions from a host language, thereby reducing the cost of defining DSLs.  \todo{Easy to define DSLs}

\end{itemize}

We are developing a comprehensive approach that has the potential to meet these design requirements well, based on defining DSLs within a language specifically designed for extensibility from within.
The approach inherently supports multiple DSLs in the same file, and
supports interoperability because the DSLs are defined by translation
into a common internal language.  One key aspect of our proposed solution, and one of the focal points of this paper, is
that we restrict the syntax of DSLs in a few key ways that greatly simplify
the mechanism without significantly decreasing expressiveness:
\begin{itemize}

\item The host language and its DSLs share a tokenization and lexing 
  strategy, standardizing conventions for identifiers, operators,
  constants, and comments.  This avoids the cost of defining lexing
  within each DSL, avoids many kinds of low-level clashes between
  languages, and makes the composed language more readable. Note that this does not limit the ability of a DSL to define new keywords and other constructs.

\item Constructs in the host language and its DSLs are
  \emph{whitespace-delimited} at the top level, according to a particular strategy that we will describe.  Various forms of parentheses 
  can also serve as delimiters.  Thus, indentation levels or
  parenthesized expressions clearly delimit blocks that are governed by a particular 
  language.  This makes the boundaries of each DSL clear to
  the programmer and the compiler, enhancing usability and guaranteeing that 
  subtle conflicts cannot arise.

\end{itemize}

Within this syntactic framework, we then propose a novel syntax extension
mechanism in which the \emph{expected type of an expression}, rather than an explicit keyword, 
%, rather than a keyword, 
signals the switch to a new DSL that parses the tokenized and lexed code within a subsequent delimited block according to a type-associated grammar. The  more common keyword-directed strategy arises as a special case of this type-directed strategy. 
We will begin with some motivating examples (\S\ref{s:motivation}), describe our approach (\S\ref{s:approach}) and implementation (\S\ref{s:implementation}), and conclude with related work (\S\ref{s:related}).

% keep discussion of type-based parsing brief - active typing for parsing (only)
% avoids conflict with Cyrus' paper

%The rest of the paper is organized as follows.  The next section
%illustrates our approach by example, discussing the components of the
%solution in more detail. Section~\ref{s:approach} outlines our
%approach, shows a wider variety of examples and discusses variations
%of our approach.  Our in-progress implementation of the proposal is
%described in Section~\ref{s:implementation}.  Section~\ref{s:related}
%compares to related work, and Section~\ref{s:conclusion} concludes.

% with a discussion of future work.

### 2013-04-11 globaldsl13/motivation.tex
Added GlobalDSL draft

\section{Motivating Examples}
\label{s:motivation}

\begin{figure}
  \centering
  \begin{lstlisting}
val hotProductDashboard = new Application
    external component twitter : Feed
        location [www.twitter.com]
    external component client : Browser
        connects to servlet
    component servlet : DashServlet
        connects to productDB, twitter
        location [intranet.nameless.com]
    component productDB : Database
        location [db.nameless.com]
    policy NamelessSecurity
        must salt [servlet.login.password]
        connect * -> servlet with HTTPS
        connect servlet -> productDB with TLS
  \end{lstlisting}
  \caption{Wyvern DSL: Architecture Specification}
  \label{f:dsl-arch}
\end{figure}

We start with a few examples meant to illustrate the expressiveness of our approach and the breadth of DSLs we plan for it to support.  The examples are presented in the proposed syntax for Wyvern, a new language being developed by our group  initially targeted toward building secure web and mobile applications. We will further clarify how these examples are parsed in Section \ref{s:approach}.

The first example, shown in Figure~\ref{f:dsl-arch}, describes the overall architecture of a ``hot product dashboard'' application.  The constructor \verb|Application| takes an argument of type \verb|ComponentArch|. Rather than being provided explicitly, a DSL for specifying the component architecture of the application is used to provide this argument because it appears in a whitespace-delimited block. The example architecture declares several components, some of which are marked \keyw{external} to indicate that they are used by this application but are not part of it directly. Component types, connectivity and properties, such as location, are declared within further-indented blocks.

The \keyw{policy} keyword introduces a security policy, which constrains the communication protocols to be used as well as
enforces secure handling of passwords.  Again, indentation delimits
the scope of the security policy.  This
%example
illustrates that one DSL may be nested within another.

\begin{figure}
  \centering
  \begin{lstlisting}
val newProds = productDB.query 
    select twHandle 
    where introduced - today < 3 months
this.prodTwt = Feed(newProds)
return prodTwt.query
    select *
    group by followed
    where count > POPULAR  
  \end{lstlisting}
  \caption{Wyvern DSL: Queries}
  \label{f:dsl-query}
\end{figure}

Figure~\ref{f:dsl-query} shows how a DSL for querying can be used from within ordinary Wyvern code.  The example shows the body of a method for computing a feed that is derived from tweets about the company's new products.  In this example, the use of a querying DSL is triggered by the use of a function, query, expecting a \verb|DBQuery|.  This type defines a particular syntax for database queries, delimited by the indented block as the reader would expect. This is similar to what can be expressed with built-in query syntax such as Microsoft's LINQ \cite{mslinq}, but in this case, defined in a user-defined type. The feed query operator may define different features, as seen on Line 5.
\begin{figure}
  \centering
  \begin{lstlisting}
html:
    head:
        title: [text "Hot Products"]
        style:
            myStylesheet
    body:
        div id="search":
            SearchBox("products")
        div id="products":
            FeedBox(servlet.hotProds())
  \end{lstlisting}
  \caption{Wyvern DSL: Presentation}
  \label{f:dsl-presentation}
\end{figure}

Finally, Figure~\ref{f:dsl-presentation} shows a DSL for rendering the hot product application in a web browser.  The DSL is based on HTML but uses our indentation-based syntax and allows Wyvern code to be integrated -- function/method calls and variables of the appropriate type can be used at any point. Literal tags are introduced by a tag name followed by any tag attributes followed by a colon.

### 2013-04-11 globaldsl13/related.tex
Added GlobalDSL draft

\section{Related Work}
\label{s:related}
The most common mechanism for creating programming language extensions is macros, as exemplified by hygienic macros in Scheme. Macros in Scheme and other Lisp-style languages are written in Lisp itself, and benefit from the regular syntax of the underlying language and the consistent use of parentheses as expression delimiters.  The syntactic nature of macros allows them to control parsing within a delimited scope, and this delimited control of parsing inspired our approach in Wyvern.  Wyvern's use of types to trigger custom parsers avoids the Lisp limitation of starting each DSL with a keyword or macro name.

Wyvern chooses whitespace delimitation rather than parenthesis delimitation because, anecdotally, many programmers find syntax in that style to be more readable.  There are several proposals for making Lisp and its dialects whitespace-delimited, including sweet-expressions \cite{sweet-expr} and SRFI-49 \cite{srfi-49} for Scheme, P4P proposal for Racket \cite{p4p-proposal}, Kernel's f-expressions \cite{fexprthesis} for Common Lisp \cite{f-expr}, and others. None of them seem to be successfully implemented and deployed within the Lisp communities, however.  This may be because Lisp programmers don't want to switch away from parentheses, yet having an alternate syntax is not by itself enough to draw non-Lisp programmers into the community.

%Therefore, our approach considers the readability of language extensions from the beginning: we attempt to devise the most comfortable and natural notations for the language extensions without bothersome extra keywords and try to keep it succinct.

Some language extensibility projects provide extension at levels of abstraction above parsing. For instance, OJ (previously, OpenJava)
\cite{Tatsubori00openjava:a} provides a macro system based on a meta-object
protocol, and Backstage Java~\cite{Palmer:2011:BJM:2048066.2048137} employs compile-time meta-programming.  Both of these systems provide macro-style rewriting of existing source code, but they provide at most limited extension of language parsing.

Other systems aim at providing more flexible forms of syntax extension compared to our whitespace-delimited approach, at the potential cost of complexity or conflicts.  Camlp4 \cite{camlp4} is a preprocessor for OCaml that offers the developer the ability to extend the concrete syntax of the language via the use of parsers and extensible grammars.  SugarJ \cite{Erdweg:2011:SLL:2048147.2048199} takes a library-centric approach which allows to syntactically extend the Java language by adding libraries. In Wyvern, the core language (that is, the \verb|Exp| sort) does not get extended directly, so conflicts cannot arise and reasoning about which DSL is applicable is much simpler.

Finally, researchers have developed a number of DSL frameworks and workbenches, such as the Meta Programming System (MPS) \cite{mps}, Spoofax \cite{KatsVisser2010}, the Intentional Domain Workbench (IDW) \cite{idw}, and Ens\={o} \cite{enso}, that provide support for extending a variety of programming languages by applying generic rules and concepts.  The Marco language \cite{lee:2012:marco} similarly provides macro definition at a level of abstraction that is largely independent of the target language.

Compared to these approaches, Wyvern focuses on making one programming language extensible with embedded DSLs, provides a common lexer and extensible parsing confined to whitespace-delimited regions, and investigates a novel type-based approach to triggering parser extensions.

%programming
%language extensions to a higher level of abstraction and creates a
%system that consumes the target language and rules for creating
%macros and produces a safe and expressive language. In contrast, our
%approach aims only at the Wyvern programming language and tries to
%enhance its extensibility.

%However, Camlp4 does not support types for the language extensions and thus is missing the additional layer of safety that our approach provides.

%While aiming for the same goal, Backstage Java  and SugarJ
% take very different approaches.

### 2013-04-22 ace/ace.tex
Updated ace draft

@@
 %%*************************************************************************
-\documentclass[10pt, conference, compsocconf]{IEEEtran}
+\documentclass{sig-alternate}
 
-\usepackage{cite}
-\renewcommand{\citepunct}{,\,} % IEEEtran wants to use ],\,[ for this but that looks dumb...
+%\usepackage{cite}
+%\renewcommand{\citepunct}{,\,} % IEEEtran wants to use ],\,[ for this but that looks dumb...
 
@@
   commentstyle=\itshape,
-  basicstyle=\ttfamily\footnotesize,
+  basicstyle=\ttfamily\scriptsize,
   morekeywords={lambda, self, assert, as},
@@
 %
-\ifCLASSINFOpdf
-  % \usepackage[pdftex]{graphicx}
-  % declare the path(s) where your graphic files are
-  % \graphicspath{{../pdf/}{../jpeg/}}
-  % and their extensions so you won't have to specify these with
-  % every instance of \includegraphics
-  % \DeclareGraphicsExtensions{.pdf,.jpeg,.png}
-\else
-  % or other class option (dvipsone, dvipdf, if not using dvips). graphicx
-  % will default to the driver specified in the system graphics.cfg if no
-  % driver is specified.
-  % \usepackage[dvips]{graphicx}
-  % declare the path(s) where your graphic files are
-  % \graphicspath{{../eps/}}
-  % and their extensions so you won't have to specify these with
-  % every instance of \includegraphics
-  % \DeclareGraphicsExtensions{.eps}
-\fi
 % *** MATH PACKAGES ***
@@
 % can use linebreaks \\ within to get better formatting as desired
-\title{\Ace: An Actively-Typed Compilation Environment\\for High-Performance Computing}
+\title{Ace: An Actively-Typed Language and Compilation Environment for High-Performance Computing}
 
@@
 % affiliations
-
-\author{\IEEEauthorblockN{Cyrus Omar, Nathan Fulton and Jonathan Aldrich}
-\IEEEauthorblockA{School of Computer Science\\
-Carnegie Mellon University\\
-Pittsburgh, PA, USA\\
-\url{http://www.acelang.org/}
-}}
+\author{\alignauthor
+Cyrus Omar and Jonathan Aldrich\\
+       \affaddr{Carnegie Mellon University, Pittsburgh, PA}\\
+       \email{\{comar,aldrich\}@cs.cmu.edu}
+}
 
@@
 \begin{abstract}
-We introduce the Ace compilation environment and demonstrate its suitability as a foundational tool for both research and practice in high-performance computing.
-Ace consists of a statically-typed programming language with {\em user-extensible semantics} specified via a compile-time meta\-language, Python.
-Users specify new primitive types and their operations by equipping type definitions, which are first-class objects in the metalanguage, with  methods that the compiler selectively invokes when type checking and translating expressions, a mechanism we call {\em active type-checking and translation (\ATT)}.
-We demonstrate the flexibility of this mechanism by implementing primitives from several widely-known languages, including the entirety of OpenCL, as libraries.
-Ace also supports more general forms of metaprogramming, and functions can be launched directly from Python with standard numeric data structures as arguments.
-Using these features, we designed a scientific simulation framework that allows users to modularly specify, compile and orchestrate the execution of parameterized families of scientific simulations on clusters of GPUs.
-This framework has been used to successfully conduct large-scale, high-performance neuroscience simulations, providing initial evidence that Ace is useful in practice today.
+Ace %, evaluates it against several design and adoption criteria, and describes several use cases as well as a realistic case study involving parallelizing neurobiological circuit simulations on clusters of GPUs.
+is an extensible statically-typed programming language embedded within the ubiquitous scripting language Python.
+Python serves as a {metalanguage}, enabling programmatic control over several aspects of language specification and compilation.
+The central innovation is a novel reification of types as first-class objects in the metalanguage, a refinement to the concept of active libraries that we call \emph{active types}.
+We demonstrate the power of this design for HPC by defining all of the primitives of the OpenCL kernel language as a library of user-defined types in Ace, then extending this core with several higher-level first-class abstractions.
+The target of translation, called the \emph{backend}, is user-defined as well, allowing us to implement the semantics of these primitives by direct translation to OpenCL, CUDA or C99.
+
+Although Ace is built around a static type discipline, it supports a novel form of generic programming based on type propagation that, together with a form of extensible local type inference, eliminates most of the syntactic overhead usually associated with statically-typed languages.
+A companion host API allows type information from \verb|numpy| arrays to directly propagate into Ace kernels, supporting seamless adoption of Ace into current scientific workflows.
+
+We validate our overall design by developing a simulation framework that uses the features of Ace to allow users to specify, compile and orchestrate the execution of ensembles of neural simulations on clusters of GPUs.
 \end{abstract}
 
-\begin{IEEEkeywords}
-programming languages, metaprogramming, type systems, extensibility, heterogeneous computing
-\end{IEEEkeywords}
+%\begin{IEEEkeywords}
+%extensible programming languages, heterogeneous computing, type systems
+%\end{IEEEkeywords}
 
@@
 % creates the second title. It will be ignored for other modes.
-\IEEEpeerreviewmaketitle
+%\IEEEpeerreviewmaketitle
 
@@
 Surveys show that scientists and engineers now spend up to 40\% of their time writing software \cite{howison2011scientific, hannay2009scientists}.
-Most of this software targets desktop hardware and about 20\% of scientists also target either local clusters or super\-computers for more numerically-intensive computations \cite{hannay2009scientists}.
-To fully harness the power of these platforms, however, these {\em professional end-user developers} \cite{segal2007some} must write parallel programs.
+Most of this software targets desktop hardware, while about 20\% of scientists also target either local clusters or super\-computers for more numerically-intensive computations \cite{hannay2009scientists}.
+To fully harness the power of these platforms, however, these {\em professional end-user developers} \cite{segal2007some} must increasingly write parallel programs.
 
-Professional end-users today generally use high-level scripting languages like MATLAB, Python, R or Perl for tasks that are not performance-sensitive, such as small-scale data analysis and plotting \cite{nguyen2010survey}.
-For portions of their analyses where these interpretted languages are too slow, they will typically call into code written in a statically-typed, low-level language like C or Fortran and use low-level parallel abstractions like pthreads or MPI \cite{4222616,basili2008understanding}.
-Unfortunately, these low-level languages and abstractions are notoriously error-prone and difficult to use, even for expert programmers.
+Professional end-users today generally use dynamically-typed scripting languages like MATLAB, Python, R or Perl for tasks that are not performance-sensitive, such as small-scale data analysis and plotting \cite{nguyen2010survey}.
+For portions of their analyses where these interpretted languages are too slow, they will typically call into code written in a statically-typed, low-level language like C or Fortran that uses low-level parallel abstractions like pthreads or MPI \cite{4222616,basili2008understanding}.
+Unfortunately, these low-level languages and abstractions are notoriously difficult to use properly, even for expert programmers.
 
-The research community has proposed dozens of novel language features and parallel programming abstractions that aim to  more equitably balance important concerns relating to performance, portability, verifiability and ease-of-use.
-Unfortunately, professional end-users are hesitant to adopt these, with many being generally skeptical regarding the practicality of new approaches that come from the research community.
+The research community has responded with dozens of language features and parallel programming abstractions that aim to  more equitably balance various concerns relating to performance, portability, verifiability and ease-of-use.
+Unfortunately, professional end-users have become skeptical that new approaches that originate in the research community are practical.
 This viewpoint was perhaps most succinctly expressed by a participant in a recent study by Basili et al.
@@
 
-Many of the barriers to adoption faced by new abstractions stem from the fact that new abstractions are often developed and distributed together with a new programming language.
-Some prominent examples where this {\em language-oriented approach} has been followed include:
+We argue that many of the barriers to adoption faced by new abstractions stem from the fact that new abstractions are too-often developed and distributed together with a new programming language.
+Some prominent cases where such a {\em language-oriented approach} \cite{languageoriented} has been followed include:
 \begin{itemize}
 \item UPC, CAF, X10 and Chapel, built around first-class support for a partitioned global address space (PGAS) programming model; 
-\item Erlang, Charm++, Occam and Axum, built around first-class support for message passing;
-\item Cilk and NESL, built around first-class support for nested data parallelism; and
-\item {\color{red} Add something about automatic parallelization of functional programs.}
-\item {\color{red} Add something about CSP like Go.}
+\item Go, Erlang, Charm++, Occam and Axum, built around first-class support for message passing; and
+\item CUDA, OpenCL, Cilk and NESL, built around first-class support for data parallelism.
 \end{itemize}
 
-There is indeed evidence suggesting that developers would prefer first-class support for a parallel abstraction over a library-based solution, when all else is held constant \cite{ppfirstclass}.
+The benefits of taking this approach are clear: first-class language, compiler and run-time support for a programming abstraction can improve performance and enable a more natural programming style.
+Empirical evidence also supports this notion that developers prefer first-class support for a parallel abstraction over a library-based solution, if other factors are held constant \cite{ppfirstclass}.
+
 Unfortunately, adopting a specialized research language comes with a host of ancillary challenges  \cite{adoptionTR,socioplt}.
-Indeed, the most widely {\em adopted} parallel programming abstractions have been those that are distributed as libraries for an existing, widely-used language, such as MapReduce, Global Arrays and MPI.
-This {\em library-oriented approach} has proven more practical because it allows developers to progressively introduce parallelism into existing programs, leverage a familiar and well-developed general-purpose programming language when not directly invoking parallel primitives, and utilize different abstractions within a single program without encountering the friction inherent to language boundaries.
+These are significant enough that programming abstractions that are distributed as libraries for a well-established language, such as MapReduce, Global Arrays and MPI, have prevailed in practice, despite the limited expressivity of library-based abstractions.
+This is balanced by several factors: developers can progressively introduce parallelism into existing programs, leverage a familiar and well-developed general-purpose programming language when not directly invoking parallel primitives, and utilize arbitrary combinations of  abstractions within a single program without needing to navigate the friction inherent to language boundaries.
 
-Ace aims to bridge these two approaches by allowing developers to define and distribute new {\em first-class language constructs} directly within user libraries, using a mechanism we introduce called {\em active typechecking and translation (\ATT)}.
-Specifically, \ATT~allows developers to  extend the semantics of a language that has a fixed, but flexible, grammar.
+Ace aims to bridge the language-oriented and library-oriented approaches by allowing developers to define and distribute a broad range of new {\em first-class} types and their operations directly within user libraries, using a mechanism called {\em active typechecking and translation (\ATT)}.
+Based on the notion of active libraries \cite{activelibraries}, \ATT~is designed to allow developers to  extend the semantics, but not the syntax, of a language.
 The Ace grammar is a nearly-complete subset of Python's  and we show that it is flexible enough to support all of the primitives of the C99 and OpenCL languages in a natural manner, when aided by a form of type inference that we describe.
@@
 $ `acec hello.py`
-Hello, compile-time world.
-Goodbye, compile-time world.
+Hello, compile-time world!
+Goodbye, compile-time world!
 $ `cat hello.cl`

### 2013-04-23 ace/ace.tex
Updated Ace intro

@@
 
-Computer-aided simulations and data analysis techniques have transformed science and engineering.
+Computer-aided simulation and data analysis techniques have transformed science and engineering.
 Surveys show that scientists and engineers now spend up to 40\% of their time writing software \cite{howison2011scientific, hannay2009scientists}.
 Most of this software targets desktop hardware, while about 20\% of scientists also target either local clusters or super\-computers for more numerically-intensive computations \cite{hannay2009scientists}.
-To fully harness the power of these platforms, however, these {\em professional end-user developers} \cite{segal2007some} must increasingly write parallel programs.
+To fully harness the power of these platforms, however, these so-called {\em professional end-user developers} \cite{segal2007some} must increasingly write parallel programs.
 
-Professional end-users today generally use dynamically-typed scripting languages like MATLAB, Python, R or Perl for tasks that are not performance-sensitive, such as small-scale data analysis and plotting \cite{nguyen2010survey}.
-For portions of their analyses where these interpretted languages are too slow, they will typically call into code written in a statically-typed, low-level language like C or Fortran that uses low-level parallel abstractions like pthreads or MPI \cite{4222616,basili2008understanding}.
-Unfortunately, these low-level languages and abstractions are notoriously difficult to use properly, even for expert programmers.
+Professional end-users today generally use dynamically-typed high-level languages like MATLAB, Python, R or Perl for tasks that are not performance-sensitive, such as small-scale data analysis and plotting \cite{nguyen2010survey}.
+For portions of their analyses where the performance overhead of dynamic type checking and automatic memory management is too high, they will typically call into code written in a statically-typed, low-level language like C or Fortran that uses low-level parallel abstractions like pthreads and MPI \cite{4222616,basili2008understanding}.
+Unfortunately, these low-level languages and abstractions are notoriously difficult to use properly because verifying code written with them remains a difficult (and often ill-specified) problem.
 
-The research community has responded with dozens of language features and parallel programming abstractions that aim to  more equitably balance various concerns relating to performance, portability, verifiability and ease-of-use.
-Unfortunately, professional end-users have become skeptical that new approaches that originate in the research community are practical.
+Researchers and domain experts often respond to these challenges by proposing novel language features that aim to strike an intermediate balance between \textbf{performance}, \textbf{verifiability\-}, \textbf{portability} and \textbf{ease-of-use}.
+Unfortunately, professional end-users rarely adopt new languages, even when they have been actively developed.
+Indeed, many end-users have become skeptical that novel approaches that originate in the research community can be  practical.
 This viewpoint was perhaps most succinctly expressed by a participant in a recent study by Basili et al.
 \cite{basili2008understanding}, who stated ``I hate MPI, I hate C++.
-[But] if I had to choose again, I would probably choose the same.'' Although this sentiment is easy to dismiss as paradoxical, we believe that it demands direct examination by researchers working to advance high-performance computing.
+[But] if I had to choose again, I would probably choose the same.'' Although it may seem paradoxical, the ubiquity of this sentiment demands direct examination by researchers proposing novel abstractions and languages for professional end-users in HPC.
 
-We argue that many of the barriers to adoption faced by new abstractions stem from the fact that new abstractions are too-often developed and distributed together with a new programming language.
-Some prominent cases where such a {\em language-oriented approach} \cite{languageoriented} has been followed include:
+We suggest three mutually-related {design criteria} that, unlike those in bold above, many languages and language-integrated abstractions have failed to adequately consider: \textbf{continuity}, \textbf{extensibility} and \textbf{interoperability}.
+These criteria encompass the intuitions that new abstractions will not be adopted in a vacuum, that programming systems must  support change, and that interacting components of an application or workflow should be able to make use of different abstractions naturally and without the possibility of conflict arising at their interface boundaries.
+%that this is far from being a superfluous issue. These criteria can influence the core design of a language or abstraction.
+
+Ace is a programming language intended to be used by professional end-users as well as researchers in areas related to high-performance computing and computational science.
+The primary contribution of this paper consists of describing several novel language mechanisms and novel combinations of existing mechanisms that Ace uses to cohesively address the design criteria above.
+We anticipate that many of these mechanisms and choices will also influence other languages in the future.
+Following a principled design methodology, we justify each design decision by appeal to this full set of design criteria and citations to relevant literature.
+These criteria and methodology constitute another contribution of this work that we hope  will lead to fewer \emph{ad hoc} designs and serve as a foundation for evaluating competing designs in HPC and related domains in the future.
+
+We begin in Section \ref{highlevel} with a high-level overview of the Ace language design and the reasoning behind the choices that were made.
+In Section \ref{usage} we provide further details by way of concrete examples employing Ace for low-level GPU programming, based on an internalization of the OpenCL programming language as an Ace library.
+Ace can be used both as a standalone language via the \verb|acec| compiler and as an interactive compilation environment.
+This latter mode of use enables just-in-time specialization and direct invocation of Ace functions from within Python scripts, as well as deep integration between the \verb|numpy| library, the OpenCL host API and the Ace compiler.
+
+In Section \ref{att}, we describe the extensible type system, which is based on the novel idea of reifying types as metalanguage objects.
+This mechanism, called \emph{active typing}, was used to implement the OpenCL library.
+The operations associated with \emph{active types} are given meaning by translation to a user-defined backend.
+We show in Section \ref{translation} how this allows us to give semantics to OpenCL constructs by translation to OpenCL itself, as well as to CUDA and C99, increasing the operational flexibility of the OpenCL standard.
+Then in Section \ref{usecases}, we  continue with several examples of higher-level parallel abstractions that can be implemented in Ace.
+In other systems, these have required deriving a new variant of an existing language or a new language, both strategies that violate at least two of the three new criteria above.
+In Ace, these exist within libraries and can safely be used within a single program.
+
+To demonstrate that Ace can be used for realistic scientific workloads, we briefly describe in Section \ref{casestudy} a case study that used Ace as the foundation for a neurobiological simulation framework capable of orchestrating the execution of thousands of realizations of a large stochastic neural circuit model on clusters of GPUs.
+We conclude in Sections \ref{related} and \ref{discussion} with related work and a discussion of some limitations of Ace at the time of writing, as well as a discussion of future work building on both Ace itself as well as the concept of active typing central to Ace.
+
+\section{The Design of Ace}\label{highlevel}
+It is now widely-acknowledged that static type systems are fundamental tools for performance and verifiability, both critical criteria in HPC, so Ace is fundamentally a statically-typed language. 
+To achieve continuity, Ace is implemented as a library in the ubiquitous scripting language Python.
+The Ace grammar is precisely that of Python, and it share's Python's packaging system, so any tools that work with Python files and packages can immediately be used with Ace.
+
+Another important tool that has become increasingly prominent is the use of code generation techniques to optimize code and orchestrate scientific workflows, so Ace includes a powerful
+
 \begin{itemize}
-\item UPC, CAF, X10 and Chapel, built around first-class support for a partitioned global address space (PGAS) programming model; 
-\item Go, Erlang, Charm++, Occam and Axum, built around first-class support for message passing; and
-\item CUDA, OpenCL, Cilk and NESL, built around first-class support for data parallelism.
+\item Types are fundamental tools for both performance and verifiability.
+\item Programmers in high-performance computing tend to be wary of abstraction mechanisms that necessitate run-time overhead.
+\item Compile-time code generation is often used to optimize code and orchestrate scientific workflows.
+\item Scripting languages, such as Python, are particularly appreciated for their ease-of-use.
+\item Library-based abstractions are more highly-adopted, but language-integrated abstractions are easier to use \cite{xxx}.
 \end{itemize}
 
-The benefits of taking this approach are clear: first-class language, compiler and run-time support for a programming abstraction can improve performance and enable a more natural programming style.
+\section{Usage of Ace}\label{usage}
+
+\section{Active Typechecking and Translation}\label{att}
+\subsection{Active Translation}\label{translation}
+\subsection{Use Cases}\label{usecases}
+\section{Case Study}\label{casestudy}
+\section{Related Work}\label{related}
+\section{Discussion}\label{discussion}
+%We argue that many of the barriers to adoption faced by new abstractions stem from the fact that new abstractions are too-often developed and distributed together with a new programming language.
+Some prominent cases where such a {\em language-oriented approach} \cite{languageoriented} has been followed include:
+%\begin{itemize}
+%\item UPC, CAF, X10 and Chapel, built around first-class support for a partitioned global address space (PGAS) programming model; 
+%\item Go, Erlang, Charm++, Occam and Axum, built around first-class support for message passing; and
+%\item CUDA, OpenCL, Cilk and NESL, built around first-class support for data parallelism.
+%\end{itemize}
+%
+%The benefits of taking this approach are clear: first-class language, compiler and run-time support for a programming abstraction can improve performance and enable a more natural programming style.
 Empirical evidence also supports this notion that developers prefer first-class support for a parallel abstraction over a library-based solution, if other factors are held constant \cite{ppfirstclass}.
-
-Unfortunately, adopting a specialized research language comes with a host of ancillary challenges  \cite{adoptionTR,socioplt}.
+%
+%Unfortunately, adopting a specialized research language comes with a host of ancillary challenges  \cite{adoptionTR,socioplt}.
 These are significant enough that programming abstractions that are distributed as libraries for a well-established language, such as MapReduce, Global Arrays and MPI, have prevailed in practice, despite the limited expressivity of library-based abstractions.
 This is balanced by several factors: developers can progressively introduce parallelism into existing programs, leverage a familiar and well-developed general-purpose programming language when not directly invoking parallel primitives, and utilize arbitrary combinations of  abstractions within a single program without needing to navigate the friction inherent to language boundaries.
-
-Ace aims to bridge the language-oriented and library-oriented approaches by allowing developers to define and distribute a broad range of new {\em first-class} types and their operations directly within user libraries, using a mechanism called {\em active typechecking and translation (\ATT)}.
+%
+%Ace aims to bridge the language-oriented and library-oriented approaches by allowing developers to define and distribute a broad range of new {\em first-class} types and their operations directly within user libraries, using a mechanism called {\em active typechecking and translation (\ATT)}.
 Based on the notion of active libraries \cite{activelibraries}, \ATT~is designed to allow developers to  extend the semantics, but not the syntax, of a language.
@@
 Notably, all of these primitives are imported via the normal library loading mechanism; they are not defined within domain-specific languages, nor as extensions for a particular compiler implementation.
-
-Although the \ATT~mechanism, a key feature of Ace, is designed primarily for use by researchers and domain experts, Ace is intended to be highly practical and ready for adoption by professional end-user developers today.
+%
+%Although the \ATT~mechanism, a key feature of Ace, is designed primarily for use by researchers and domain experts, Ace is intended to be highly practical and ready for adoption by professional end-user developers today.
 It is based on the Python programming language, which enjoys broad adoption in this community.
 Indeed, the language itself is implemented entirely as a Python library, and many existing tools, such as editor modes, documentation tools and style checkers, can be used without modification.
-Ace functions are generated programmatically and can be called directly from Python, with numeric arrays as arguments, so developers can integrate Ace immediately within their existing workflows for portions of code that may benefit from static compilation and support for parallelism.
-We demonstrate the practicality of Ace in Section \ref{clegans} by describing an Ace-based framework that supports specifying and orchestrating families of large-scale neurobiological circuit simulations on clusters of GPUs.
+%Ace functions are generated programmatically and can be called directly from Python, with numeric arrays as arguments, so developers can integrate Ace immediately within their existing workflows for portions of code that may benefit from static compilation and support for parallelism.
+%We demonstrate the practicality of Ace in Section \ref{clegans} by describing an Ace-based framework that supports specifying and orchestrating families of large-scale neurobiological circuit simulations on clusters of GPUs.
 

### 2013-04-23 ace/ace.tex
Updated abstract per PLAID group suggestions

@@
 \definecolor{mauve}{rgb}{0.58,0,0.82}
+\definecolor{light-gray}{gray}{0.75}
 \usepackage{listings}
@@
   numbers=left,
-  numberstyle=\footnotesize\color{gray}\textsf,
+  numberstyle=\scriptsize\color{light-gray}\textsf,
   xleftmargin=2em,
@@
 \begin{abstract}
-Ace %, evaluates it against several design and adoption criteria, and describes several use cases as well as a realistic case study involving parallelizing neurobiological circuit simulations on clusters of GPUs.
-is an extensible statically-typed programming language embedded within the ubiquitous scripting language Python.
+Researchers developing languages and abstractions for high-performance computing must consider a number of design criteria, including performance, verifiability, portability and ease-of-use.
+Despite the deficiencies of today's tools and the availability of apparently superior options, end-users have been reluctant to adopt new language-based abstractions.
+We argue that this can be largely attributed to a failure to consider three critical criteria: continuity, extensibility and interoperability.
+This paper introduces Ace, a language that incorporates several novel mechanisms and makes principled design choices in order to satisfy this more comprehensive set of criteria.
+Ace builds upon existing standards in HPC, particularly Python and OpenCL, and it can be modularly extended with abstractions drawn from several previously disparate paradigms.
+The core innovation supporting this is a novel reification of types as first-class objects at compile-time, a refinement to the concept of active libraries that we call \emph{active types}.
+We validate our overall design with a case study of a simulation framework enabling the modular specification and efficient execution of ensembles of neural simulations across clusters of GPUs.
+
+%in the context of the applications and architectures that the language may need to target.
+%
+%Programming languages targeted toward end-users in high-performance computing must consider a number of issues in addition to performance.
+Typically, researchers have focused verifiability, portability and ease-of-use, and the right balance often depends on the particular uses that they are put toward.
+%
+%Ace %, evaluates it against several design and adoption criteria, and describes several use cases as well as a realistic case study involving parallelizing neurobiological circuit simulations on clusters of GPUs.
+%is an extensible statically-typed programming language embedded within the ubiquitous scripting language Python.
 Python serves as a {metalanguage}, enabling programmatic control over several aspects of language specification and compilation.
-The central innovation is a novel reification of types as first-class objects in the metalanguage, a refinement to the concept of active libraries that we call \emph{active types}.
-We demonstrate the power of this design for HPC by defining all of the primitives of the OpenCL kernel language as a library of user-defined types in Ace, then extending this core with several higher-level first-class abstractions.
+%We demonstrate the power of this design for HPC by defining all of the primitives of the OpenCL kernel language as a library of user-defined types in Ace, then extending this core with several higher-level first-class abstractions.
 The target of translation, called the \emph{backend}, is user-defined as well, allowing us to implement the semantics of these primitives by direct translation to OpenCL, CUDA or C99.
-
-Although Ace is built around a static type discipline, it supports a novel form of generic programming based on type propagation that, together with a form of extensible local type inference, eliminates most of the syntactic overhead usually associated with statically-typed languages.
+%
+%Although Ace is built around a static type discipline, it supports a novel form of generic programming based on type propagation that, together with a form of extensible local type inference, eliminates most of the syntactic overhead usually associated with statically-typed languages.
 A companion host API allows type information from \verb|numpy| arrays to directly propagate into Ace kernels, supporting seamless adoption of Ace into current scientific workflows.
-
-We validate our overall design by developing a simulation framework that uses the features of Ace to allow users to specify, compile and orchestrate the execution of ensembles of neural simulations on clusters of GPUs.
+%
 \end{abstract}
@@
 Professional end-users today generally use dynamically-typed high-level languages like MATLAB, Python, R or Perl for tasks that are not performance-sensitive, such as small-scale data analysis and plotting \cite{nguyen2010survey}.
-For portions of their analyses where the performance overhead of dynamic type checking and automatic memory management is too high, they will typically call into code written in a statically-typed, low-level language like C or Fortran that uses low-level parallel abstractions like pthreads and MPI \cite{4222616,basili2008understanding}.
-Unfortunately, these low-level languages and abstractions are notoriously difficult to use properly because verifying code written with them remains a difficult (and often ill-specified) problem.
+For portions of their analyses where the performance overhead of dynamic type checking and automatic memory management is too high, they will typically call into code written in a statically-typed, low-level language, most commonly C or Fortran, that uses low-level parallel abstractions like pthreads and MPI \cite{4222616,basili2008understanding}.
+Unfortunately, these low-level languages and abstractions are notoriously difficult to use and code correctness cannot be automatically verified in general.
 
 Researchers and domain experts often respond to these challenges by proposing novel language features that aim to strike an intermediate balance between \textbf{performance}, \textbf{verifiability\-}, \textbf{portability} and \textbf{ease-of-use}.
-Unfortunately, professional end-users rarely adopt new languages, even when they have been actively developed.
+Unfortunately, professional end-users rarely adopt new languages.
 Indeed, many end-users have become skeptical that novel approaches that originate in the research community can be  practical.
 This viewpoint was perhaps most succinctly expressed by a participant in a recent study by Basili et al.
-\cite{basili2008understanding}, who stated ``I hate MPI, I hate C++.
-[But] if I had to choose again, I would probably choose the same.'' Although it may seem paradoxical, the ubiquity of this sentiment demands direct examination by researchers proposing novel abstractions and languages for professional end-users in HPC.
+\cite{basili2008understanding} who stated ``I hate MPI, I hate C++.
+[But] if I had to choose again, I would probably choose the same.'' Although it may seem paradoxical, the ubiquity of this sentiment demands direct examination by researchers proposing novel abstractions and languages for eventual use by professional end-users in HPC.
 

### 2013-04-24 ace/ace.tex
Updated ace abstract and intro

@@
 Researchers developing languages and abstractions for high-performance computing must consider a number of design criteria, including performance, verifiability, portability and ease-of-use.
-Despite the deficiencies of today's tools and the availability of apparently superior options, end-users have been reluctant to adopt new language-based abstractions.
-We argue that this can be largely attributed to a failure to consider three critical criteria: continuity, extensibility and interoperability.
-This paper introduces Ace, a language that incorporates several novel mechanisms and makes principled design choices in order to satisfy this more comprehensive set of criteria.
-Ace builds upon existing standards in HPC, particularly Python and OpenCL, and it can be modularly extended with abstractions drawn from several previously disparate paradigms.
-The core innovation supporting this is a novel reification of types as first-class objects at compile-time, a refinement to the concept of active libraries that we call \emph{active types}.
-We validate our overall design with a case study of a simulation framework enabling the modular specification and efficient execution of ensembles of neural simulations across clusters of GPUs.
+Despite the deficiencies of legacy tools and the availability of seemingly superior options, end-users have been reluctant to adopt new language-based abstractions.
+We argue that this can be largely attributed to a failure to consider three additional criteria: continuity, extensibility\- and interoperability.
+This paper introduces Ace, a language that aims to satisfy this more comprehensive set of design criteria.
+To do so, Ace introduces several novel compile-time mechanisms, makes principled design choices, and builds upon existing standards in HPC, particularly Python and OpenCL.
+OpenCL support is not built into the language, but rather implemented atop an extensibility mechanism that also admits abstractions drawn from several other seemingly disparate paradigms.
+The core innovation underlying this and several other features of Ace is a novel reification of types as first-class objects at compile-time, a refinement to the concept of active libraries that we call \emph{active types}.
+We validate our overall design by considering a case study of a simulation framework enabling the modular specification and efficient execution of ensembles of neural simulations across clusters of GPUs.
 
@@
 For portions of their analyses where the performance overhead of dynamic type checking and automatic memory management is too high, they will typically call into code written in a statically-typed, low-level language, most commonly C or Fortran, that uses low-level parallel abstractions like pthreads and MPI \cite{4222616,basili2008understanding}.
-Unfortunately, these low-level languages and abstractions are notoriously difficult to use and code correctness cannot be automatically verified in general.
+Unfortunately, these low-level languages and abstractions are notoriously difficult to use and automatic verification is intractable in general.
 
@@
 
-Ace is a programming language intended to be used by professional end-users as well as researchers in areas related to high-performance computing and computational science.
+In this paper, we introduce Ace, a programming language targeting professional end-users as well as researchers across high-performance computing and related domains.
+Ace has followed a principled design methodology guided by this more comprehensive set of design criteria in order to avoid many of the issues that have hindered previous language-based approaches in high-performance computing.
+These criteria, this design methodology, and the novel mechanisms and designs developed to satisfy these criteria constitute the generalizable contributions of this paper.
+We also hope that Ace itself will be useful to the HPC community\footnote{\small Ace is openly available at http://acelang.org/.}, and we describe several use cases and an initial case study as a means to preliminarily validate it's utility.
+
+ We begin in Section \ref{usage} with simple examples that show how Ace can be used for low-level GPU programming and introduce the fundamental decisions made in its design.
+We discuss the motivations, based on the criteria above, behind key design decisions, including the use of static typing, a fixed syntax based on Python, a novel type propagation and type inference scheme, and an explicit phase separation between compile-time and run-time logic.
+
+The examples in Section \ref{usage} are based on an internalization of the OpenCL kernel programming language as a library.
+This library is built atop a novel extensibility mechanism that we call \emph{active typechecking and translation} (AT\&T) and detail in Section \ref{att}.
+This mechanism relies on the core idea of representing types as metalanguage objects.
+We refer to such types as \emph{active types} because they are a refinement to the concept of \emph{active libraries} (cf.
+Section \ref{related}).
+Unlike prior approaches to extensibility that take a language-global view, hindering  interoperability, the AT\&T mechanism guarantees that extensions are composable by construction.
+
+%Despite its simplicity and safety, this mechanism is flexible enough to define the entirety of the OpenCL kernel language (which is derived directly from C99) and, combined with Ace's type propagation scheme, substantially enhance it's ease-of-use.
+To further demonstrate the flexibility of this mechanism, we continue in Section \ref{usecases} by outlining examples of other, higher-level parallel abstractions, including global address spaces, message passing and functional data parallelism, that can be cleanly implemented using AT\&T.
+First-class support for each of these has required a new language in the past.
+AT\&T enables compile-time logic to be coupled with user-defined types, and thus all of these first-class abstractions can safely be used within a single program or workflow.
+
+ Ace can be used as a standalone language via the \verb|acec| compiler and also as an interactive compilation environment.
+This latter mode of use supports just-in-time specialization and direct invocation of Ace functions from within Python scripts, as well as deep integration between the widely-used \verb|numpy| library (including its internal type system),  OpenCL's host API (by a mechanism also available to other host APIs, such as CUDA's) and the Ace compiler itself.
+We describe this in Section \ref{compenv}.
+Achieving this deep integration makes further use of Ace's representation of types.
+
+To demonstrate the utility of Ace for realistic scientific workloads, we briefly describe in Section \ref{casestudy} a case study that uses Ace as a compilation environment supporting a scientific simulation framework capable of orchestrating the execution of thousands of realizations of a large stochastic neurobiological circuit model on clusters of GPUs.
+
+We conclude in Sections \ref{related} and \ref{discussion} with related work and a discussion of some limitations of Ace at the time of writing, as well as a discussion of future work that builds upon the concept of active typing central to Ace.
+
+% a programming language intended to be used by professional end-users as well as researchers in areas related to high-performance computing and computational science.
 The primary contribution of this paper consists of describing several novel language mechanisms and novel combinations of existing mechanisms that Ace uses to cohesively address the design criteria above.
 We anticipate that many of these mechanisms and choices will also influence other languages in the future.
-Following a principled design methodology, we justify each design decision by appeal to this full set of design criteria and citations to relevant literature.
+%Following a principled design methodology, we justify each design decision by appeal to this full set of design criteria and citations to relevant literature.
 These criteria and methodology constitute another contribution of this work that we hope  will lead to fewer \emph{ad hoc} designs and serve as a foundation for evaluating competing designs in HPC and related domains in the future.
 
-We begin in Section \ref{highlevel} with a high-level overview of the Ace language design and the reasoning behind the choices that were made.
-In Section \ref{usage} we provide further details by way of concrete examples employing Ace for low-level GPU programming, based on an internalization of the OpenCL programming language as an Ace library.
-Ace can be used both as a standalone language via the \verb|acec| compiler and as an interactive compilation environment.
-This latter mode of use enables just-in-time specialization and direct invocation of Ace functions from within Python scripts, as well as deep integration between the \verb|numpy| library, the OpenCL host API and the Ace compiler.
-
-In Section \ref{att}, we describe the extensible type system, which is based on the novel idea of reifying types as metalanguage objects.
-This mechanism, called \emph{active typing}, was used to implement the OpenCL library.
-The operations associated with \emph{active types} are given meaning by translation to a user-defined backend.
-We show in Section \ref{translation} how this allows us to give semantics to OpenCL constructs by translation to OpenCL itself, as well as to CUDA and C99, increasing the operational flexibility of the OpenCL standard.
-Then in Section \ref{usecases}, we  continue with several examples of higher-level parallel abstractions that can be implemented in Ace.
-In other systems, these have required deriving a new variant of an existing language or a new language, both strategies that violate at least two of the three new criteria above.
-In Ace, these exist within libraries and can safely be used within a single program.
-
-To demonstrate that Ace can be used for realistic scientific workloads, we briefly describe in Section \ref{casestudy} a case study that used Ace as the foundation for a neurobiological simulation framework capable of orchestrating the execution of thousands of realizations of a large stochastic neural circuit model on clusters of GPUs.
-We conclude in Sections \ref{related} and \ref{discussion} with related work and a discussion of some limitations of Ace at the time of writing, as well as a discussion of future work building on both Ace itself as well as the concept of active typing central to Ace.
-
-\section{The Design of Ace}\label{highlevel}
+\section{Design and Usage}\label{usage}
 It is now widely-acknowledged that static type systems are fundamental tools for performance and verifiability, both critical criteria in HPC, so Ace is fundamentally a statically-typed language. 
@@
 
-\section{Usage of Ace}\label{usage}
-
-\section{Active Typechecking and Translation}\label{att}
+\section{Active Typechecking and\\Translation (AT\&T)}\label{att}
 \subsection{Active Translation}\label{translation}
 \subsection{Use Cases}\label{usecases}
+\section{Ace: A Compilation Environment}\label{compenv}
 \section{Case Study}\label{casestudy}

### 2013-04-24 ace/ace.tex
Updated abstract

@@
 To do so, Ace introduces several novel compile-time mechanisms, makes principled design choices, and builds upon existing standards in HPC, particularly Python and OpenCL.
-OpenCL support is not built into the language, but rather implemented atop an extensibility mechanism that also admits abstractions drawn from several other seemingly disparate paradigms.
-The core innovation underlying this and several other features of Ace is a novel reification of types as first-class objects at compile-time, a refinement to the concept of active libraries that we call \emph{active types}.
+OpenCL support, rather than being built into the language, is implemented atop an extensibility mechanism that also admits abstractions drawn from other seemingly disparate paradigms.
+The core innovation underlying this and other features of Ace is a novel reification of types as first-class objects at compile-time, representing a refinement to the concept of active libraries that we call \emph{active types}.
 We validate our overall design by considering a case study of a simulation framework enabling the modular specification and efficient execution of ensembles of neural simulations across clusters of GPUs.

### 2013-04-24 ace/ace.tex
Updated intro

@@
 This mechanism relies on the core idea of representing types as metalanguage objects.
-We refer to such types as \emph{active types} because they are a refinement to the concept of \emph{active libraries} (cf.
-Section \ref{related}).
-Unlike prior approaches to extensibility that take a language-global view, hindering  interoperability, the AT\&T mechanism guarantees that extensions are composable by construction.
+We refer to such types as \emph{active types} by analogy to \emph{active libraries} \cite{activelibraries} (see Section \ref{related}).
+Unlike prior approaches to extensibility that take a language-global view, which impacts interoperability, the AT\&T mechanism features a guarantee that extensions are composable by construction.
 
 %Despite its simplicity and safety, this mechanism is flexible enough to define the entirety of the OpenCL kernel language (which is derived directly from C99) and, combined with Ace's type propagation scheme, substantially enhance it's ease-of-use.
-To further demonstrate the flexibility of this mechanism, we continue in Section \ref{usecases} by outlining examples of other, higher-level parallel abstractions, including global address spaces, message passing and functional data parallelism, that can be cleanly implemented using AT\&T.
+To demonstrate the flexibility of this mechanism beyond OpenCL, we continue in Section \ref{usecases} by outlining examples of other, higher-level parallel abstractions that can be cleanly implemented using AT\&T, including global address spaces, message passing and functional data parallelism.
 First-class support for each of these has required a new language in the past.
-AT\&T enables compile-time logic to be coupled with user-defined types, and thus all of these first-class abstractions can safely be used within a single program or workflow.
+AT\&T allows compile-time logic to be safely included within libraries, and thus these first-class abstractions can coexist within a single program or workflow.
 
 Ace can be used as a standalone language via the \verb|acec| compiler and also as an interactive compilation environment.
-This latter mode of use supports just-in-time specialization and direct invocation of Ace functions from within Python scripts, as well as deep integration between the widely-used \verb|numpy| library (including its internal type system),  OpenCL's host API (by a mechanism also available to other host APIs, such as CUDA's) and the Ace compiler itself.
-We describe this in Section \ref{compenv}.
-Achieving this deep integration makes further use of Ace's representation of types.
-
-To demonstrate the utility of Ace for realistic scientific workloads, we briefly describe in Section \ref{casestudy} a case study that uses Ace as a compilation environment supporting a scientific simulation framework capable of orchestrating the execution of thousands of realizations of a large stochastic neurobiological circuit model on clusters of GPUs.
-
-We conclude in Sections \ref{related} and \ref{discussion} with related work and a discussion of some limitations of Ace at the time of writing, as well as a discussion of future work that builds upon the concept of active typing central to Ace.
+This mode of use, described in Section \ref{compenv}, uses just-in-time specialization to deeply integrate the widely-adopted \verb|numpy| library (including its internal type system) and OpenCL's host API (by a mechanism also available to other host APIs, such as CUDA's) with the Ace compiler itself to enable the direct invocation of Ace functions from within Python scripts with minimal syntactic overhead.
+Achieving this integration also makes use of Ace's representation of types.
+
+To demonstrate the utility of Ace for scientific workloads, we describe in Section \ref{casestudy} a case study where Ace was used as the underlying compilation environment for a scientific simulation framework.
+This framework was used to  specify and efficiently execute thousands of realizations of a large stochastic neurobiological circuit model on clusters of GPUs.
+
+We conclude in Sections \ref{related} and \ref{discussion} with related work and a discussion of the limitations of Ace at the time of writing, as well as a discussion of planned future work building upon the concept of active typing.
 

### 2013-04-25 ace/ace.tex
high-level design

@@
 These criteria, this design methodology, and the novel mechanisms and designs developed to satisfy these criteria constitute the generalizable contributions of this paper.
-We also hope that Ace itself will be useful to the HPC community\footnote{\small Ace is openly available at http://acelang.org/.}, and we describe several use cases and an initial case study as a means to preliminarily validate it's utility.
+We also hope that Ace itself will be useful to the HPC community\footnote{\small Ace is openly available at http://acelang.org/.}, and we describe several use cases and an initial case study in order to preliminarily validate it's utility.
 
@@
 This mechanism relies on the core idea of representing types as metalanguage objects.
-We refer to such types as \emph{active types} by analogy to \emph{active libraries} \cite{activelibraries} (see Section \ref{related}).
-Unlike prior approaches to extensibility that take a language-global view, which impacts interoperability, the AT\&T mechanism features a guarantee that extensions are composable by construction.
+We refer to such types as \emph{active types} by relation to \emph{active libraries} \cite{activelibraries} (see Section \ref{related}).
+Unlike prior approaches to extensibility that take a language-global view, which hinders interoperability, the AT\&T mechanism guarantees that extensions are composable by construction by limiting their scope to expressions having a type defined by that extension.
 
@@
 First-class support for each of these has required a new language in the past.
-AT\&T allows compile-time logic to be safely included within libraries, and thus these first-class abstractions can coexist within a single program or workflow.
+AT\&T allows compile-time logic to be safely included within libraries, and thus these first-class abstractions can be  used naturally within a single program or workflow.
 
 Ace can be used as a standalone language via the \verb|acec| compiler and also as an interactive compilation environment.
-This mode of use, described in Section \ref{compenv}, uses just-in-time specialization to deeply integrate the widely-adopted \verb|numpy| library (including its internal type system) and OpenCL's host API (by a mechanism also available to other host APIs, such as CUDA's) with the Ace compiler itself to enable the direct invocation of Ace functions from within Python scripts with minimal syntactic overhead.
-Achieving this integration also makes use of Ace's representation of types.
-
-To demonstrate the utility of Ace for scientific workloads, we describe in Section \ref{casestudy} a case study where Ace was used as the underlying compilation environment for a scientific simulation framework.
-This framework was used to  specify and efficiently execute thousands of realizations of a large stochastic neurobiological circuit model on clusters of GPUs.
-
-We conclude in Sections \ref{related} and \ref{discussion} with related work and a discussion of the limitations of Ace at the time of writing, as well as a discussion of planned future work building upon the concept of active typing.
+This mode of use, described in Section \ref{compenv}, uses just-in-time specialization to  integrate the widely-adopted \verb|numpy| library (including its internal type system) and OpenCL's host API (by a mechanism also available to other host APIs, such as CUDA's) with the Ace compiler itself to enable the direct invocation of Ace functions from within Python scripts with minimal overhead.
+Achieving this deep level of integration makes use of Ace's representation of types as well.
+
+To demonstrate the utility of Ace for scientific workloads, we describe in Section \ref{casestudy} a case study where Ace was used in this mode to develop a scientific simulation framework.
+This framework has been used to  specify and efficiently execute thousands of realizations of a large stochastic neural circuit model on clusters of GPUs, achieving the same performance as raw OpenCL code while being more modular and concise than was feasible before.
+
+We conclude in Sections \ref{related} and \ref{discussion} with related work and a discussion of the limitations of Ace at the time of writing, as well as a discussion of planned future work further validating Ace and building upon the concept of active typing.
 
@@
 
-\section{Design and Usage}\label{usage}
-It is now widely-acknowledged that static type systems are fundamental tools for performance and verifiability, both critical criteria in HPC, so Ace is fundamentally a statically-typed language. 
-To achieve continuity, Ace is implemented as a library in the ubiquitous scripting language Python.
-The Ace grammar is precisely that of Python, and it share's Python's packaging system, so any tools that work with Python files and packages can immediately be used with Ace.
-
-Another important tool that has become increasingly prominent is the use of code generation techniques to optimize code and orchestrate scientific workflows, so Ace includes a powerful
-
-\begin{itemize}
-\item Types are fundamental tools for both performance and verifiability.
-\item Programmers in high-performance computing tend to be wary of abstraction mechanisms that necessitate run-time overhead.
-\item Compile-time code generation is often used to optimize code and orchestrate scientific workflows.
-\item Scripting languages, such as Python, are particularly appreciated for their ease-of-use.
-\item Library-based abstractions are more highly-adopted, but language-integrated abstractions are easier to use \cite{xxx}.
-\end{itemize}
-
-\section{Active Typechecking and\\Translation (AT\&T)}\label{att}
-\subsection{Active Translation}\label{translation}
-\subsection{Use Cases}\label{usecases}
-\section{Ace: A Compilation Environment}\label{compenv}
-\section{Case Study}\label{casestudy}
-\section{Related Work}\label{related}
-\section{Discussion}\label{discussion}
-%We argue that many of the barriers to adoption faced by new abstractions stem from the fact that new abstractions are too-often developed and distributed together with a new programming language.
-Some prominent cases where such a {\em language-oriented approach} \cite{languageoriented} has been followed include:
-%\begin{itemize}
-%\item UPC, CAF, X10 and Chapel, built around first-class support for a partitioned global address space (PGAS) programming model; 
-%\item Go, Erlang, Charm++, Occam and Axum, built around first-class support for message passing; and
-%\item CUDA, OpenCL, Cilk and NESL, built around first-class support for data parallelism.
-%\end{itemize}
-%
-%The benefits of taking this approach are clear: first-class language, compiler and run-time support for a programming abstraction can improve performance and enable a more natural programming style.
-Empirical evidence also supports this notion that developers prefer first-class support for a parallel abstraction over a library-based solution, if other factors are held constant \cite{ppfirstclass}.
-%
-%Unfortunately, adopting a specialized research language comes with a host of ancillary challenges  \cite{adoptionTR,socioplt}.
-These are significant enough that programming abstractions that are distributed as libraries for a well-established language, such as MapReduce, Global Arrays and MPI, have prevailed in practice, despite the limited expressivity of library-based abstractions.
-This is balanced by several factors: developers can progressively introduce parallelism into existing programs, leverage a familiar and well-developed general-purpose programming language when not directly invoking parallel primitives, and utilize arbitrary combinations of  abstractions within a single program without needing to navigate the friction inherent to language boundaries.
-%
-%Ace aims to bridge the language-oriented and library-oriented approaches by allowing developers to define and distribute a broad range of new {\em first-class} types and their operations directly within user libraries, using a mechanism called {\em active typechecking and translation (\ATT)}.
-Based on the notion of active libraries \cite{activelibraries}, \ATT~is designed to allow developers to  extend the semantics, but not the syntax, of a language.
-The Ace grammar is a nearly-complete subset of Python's  and we show that it is flexible enough to support all of the primitives of the C99 and OpenCL languages in a natural manner, when aided by a form of type inference that we describe.
-We further supplement these primitives with higher-level user-defined primitives, examples of which we discuss in Section \ref{abstractions}.
-Notably, all of these primitives are imported via the normal library loading mechanism; they are not defined within domain-specific languages, nor as extensions for a particular compiler implementation.
-%
-%Although the \ATT~mechanism, a key feature of Ace, is designed primarily for use by researchers and domain experts, Ace is intended to be highly practical and ready for adoption by professional end-user developers today.
-It is based on the Python programming language, which enjoys broad adoption in this community.
-Indeed, the language itself is implemented entirely as a Python library, and many existing tools, such as editor modes, documentation tools and style checkers, can be used without modification.
-%Ace functions are generated programmatically and can be called directly from Python, with numeric arrays as arguments, so developers can integrate Ace immediately within their existing workflows for portions of code that may benefit from static compilation and support for parallelism.
-%We demonstrate the practicality of Ace in Section \ref{clegans} by describing an Ace-based framework that supports specifying and orchestrating families of large-scale neurobiological circuit simulations on clusters of GPUs.
-
+\section{Language Design and Usage}\label{usage}
+\subsection{High-Level Design}
+Static type systems are powerful tools for programming language design and implementation.
+By tracking the type of a value statically, a typechecker can verify the absence of many kinds of errors over all inputs.
+This simplifies and increases the performance of the run-time system, as errors need not be detected dynamically using tag checks and other kinds of assertions.
+Many parallel programming abstractions are defined in terms of, or benefit from, a type system that enforces a communication protocol, ensures the consistency of data and simplifies the dynamics of the run-time system (see Section \ref{usecases} for examples).
+Because \textbf{verifiability} and \textbf{performance} are key criteria and static typing is a core technique, Ace is fundamentally statically-typed.
+
+It is legitimate to ask, however, why dynamically-typed languages are so widely-used in HPC.
+One reason has to do with the criteria of \textbf{ease-of-use}.
+Statically-typed languages, particularly those that HPC programmers are exposed to, can be verbose, requiring explicit type annotations on every function and variable declaration,  separate header files, and other declarations that exist to assist the compiler.
+Cordy identifies the principle of \emph{conciseness} with elimination of
+redundancy and the availability of reasonable defaults \cite{cordy1992hints}.
+The high-level languages used in HPC have, consistent with this principle, avoided most of this overhead.
+Ace was originally conceived to explore the question: does this require run-time mechanisms, or can one express a statically-typed, low-level language at least as powerful as C without the overhead that is typically associated with it?
+
+Rather than designing a new syntax, or modifying the syntax of C, we chose to directly use, without modification, an existing language, Python's, syntax.
+This was not an arbitrary choice but rather a key mechanism by which Ace achieves \textbf{continuity}.
+Python is amongst the most widely-adopted languages in the world, and enjoys similar  ubiquity within HPC specifically \cite{pythonhpc}.
+By directly adopting Python's syntax, Ace's syntax is immediately familiar and acceptable to a significant portion of it's intended audience.
+Moreover, any tools that handle Python source code, including parsers, editors, style checkers and documentation generators, can be used without modification for handling Ace code.
+
+\subsection{Example 1: Hello, World!}
 \begin{codelisting}
@@
 
-\section{Structure and Usage}
-%Ace is an {\em actively-typed, language-integrated compilation environment}. Let us begin by defining each of these terms. 
-%A {\em compilation environment} is a system that allows developers to programmatically control aspects of compilation directly.
-The language used to do so is called the {\em metalanguage} while the grammar that programs themselves are written in is called the {\em source grammar}.
-The metalanguage of Ace is Python and the source grammar is a significant subset of Python's own grammar.
-Programs and metalanguage code can be interleaved within the same file, so Ace is {\em language-integrated}.
-Ace programs are statically-typed, and types themselves are objects in the metalanguage.
-They can be equipped by developers with methods that control how operators associated with them are typechecked and compiled.
-Because of the relationship between this mechanism and the idea of {\em active libraries} \cite{activelibraries}, discussed further in Section \ref{related}, we call them {\em active types}, and thus Ace is {\em actively-typed}.
-\subsection{Example 1: Hello, World!}
 To demonstrate the somewhat unconventional structure of Ace programs and libraries, we begin in Listing \ref{hello} with a variant of the canonical ``Hello, World!'' example written using the \verb|OpenCL| module.
@@
 
+\section{Active Typechecking and\\Translation (AT\&T)}\label{att}
+\subsection{Active Translation}\label{translation}
+\subsection{Use Cases}\label{usecases}
+\section{Ace: A Compilation Environment}\label{compenv}
+\section{Case Study}\label{casestudy}
+\section{Related Work}\label{related}
+\section{Discussion}\label{discussion}
+%We argue that many of the barriers to adoption faced by new abstractions stem from the fact that new abstractions are too-often developed and distributed together with a new programming language.
+Some prominent cases where such a {\em language-oriented approach} \cite{languageoriented} has been followed include:
+%\begin{itemize}
+%\item UPC, CAF, X10 and Chapel, built around first-class support for a partitioned global address space (PGAS) programming model; 
+%\item Go, Erlang, Charm++, Occam and Axum, built around first-class support for message passing; and
+%\item CUDA, OpenCL, Cilk and NESL, built around first-class support for data parallelism.
+%\end{itemize}
+%
+%The benefits of taking this approach are clear: first-class language, compiler and run-time support for a programming abstraction can improve performance and enable a more natural programming style.
+Empirical evidence also supports this notion that developers prefer first-class support for a parallel abstraction over a library-based solution, if other factors are held constant \cite{ppfirstclass}.
+%
+%Unfortunately, adopting a specialized research language comes with a host of ancillary challenges  \cite{adoptionTR,socioplt}.
+These are significant enough that programming abstractions that are distributed as libraries for a well-established language, such as MapReduce, Global Arrays and MPI, have prevailed in practice, despite the limited expressivity of library-based abstractions.
+This is balanced by several factors: developers can progressively introduce parallelism into existing programs, leverage a familiar and well-developed general-purpose programming language when not directly invoking parallel primitives, and utilize arbitrary combinations of  abstractions within a single program without needing to navigate the friction inherent to language boundaries.
+%
+%Ace aims to bridge the language-oriented and library-oriented approaches by allowing developers to define and distribute a broad range of new {\em first-class} types and their operations directly within user libraries, using a mechanism called {\em active typechecking and translation (\ATT)}.
+Based on the notion of active libraries \cite{activelibraries}, \ATT~is designed to allow developers to  extend the semantics, but not the syntax, of a language.
+The Ace grammar is a nearly-complete subset of Python's  and we show that it is flexible enough to support all of the primitives of the C99 and OpenCL languages in a natural manner, when aided by a form of type inference that we describe.
+We further supplement these primitives with higher-level user-defined primitives, examples of which we discuss in Section \ref{abstractions}.
+Notably, all of these primitives are imported via the normal library loading mechanism; they are not defined within domain-specific languages, nor as extensions for a particular compiler implementation.
+%
+%Although the \ATT~mechanism, a key feature of Ace, is designed primarily for use by researchers and domain experts, Ace is intended to be highly practical and ready for adoption by professional end-user developers today.
+It is based on the Python programming language, which enjoys broad adoption in this community.
+Indeed, the language itself is implemented entirely as a Python library, and many existing tools, such as editor modes, documentation tools and style checkers, can be used without modification.
+%Ace functions are generated programmatically and can be called directly from Python, with numeric arrays as arguments, so developers can integrate Ace immediately within their existing workflows for portions of code that may benefit from static compilation and support for parallelism.
+%We demonstrate the practicality of Ace in Section \ref{clegans} by describing an Ace-based framework that supports specifying and orchestrating families of large-scale neurobiological circuit simulations on clusters of GPUs.
+
+\section{Structure and Usage}
+%Ace is an {\em actively-typed, language-integrated compilation environment}. Let us begin by defining each of these terms. 
+%A {\em compilation environment} is a system that allows developers to programmatically control aspects of compilation directly.
+The language used to do so is called the {\em metalanguage} while the grammar that programs themselves are written in is called the {\em source grammar}.
+The metalanguage of Ace is Python and the source grammar is a significant subset of Python's own grammar.
+Programs and metalanguage code can be interleaved within the same file, so Ace is {\em language-integrated}.
+Ace programs are statically-typed, and types themselves are objects in the metalanguage.
+They can be equipped by developers with methods that control how operators associated with them are typechecked and compiled.
+Because of the relationship between this mechanism and the idea of {\em active libraries} \cite{activelibraries}, discussed further in Section \ref{related}, we call them {\em active types}, and thus Ace is {\em actively-typed}.
 \section{Ace for Researchers}

### 2013-04-26 ace/ace.tex
Updated high-level goals

@@
  We begin in Section \ref{usage} with simple examples that show how Ace can be used for low-level GPU programming and introduce the fundamental decisions made in its design.
-We discuss the motivations, based on the criteria above, behind key design decisions, including the use of static typing, a fixed syntax based on Python, a novel type propagation and type inference scheme, and an explicit phase separation between compile-time and run-time logic.
-
-The examples in Section \ref{usage} are based on an internalization of the OpenCL kernel programming language as a library.
+We discuss the motivations, based on the criteria above, behind key design decisions, including the use of static typing, a fixed syntax based on Python, a novel type propagation and inference scheme, and an explicit phase separation between compile-time and run-time logic.
+Each of these can be seen in Listings 1 and 2, which we will discuss further below.
+
+All examples in Section \ref{usage} are based on an internalization of the OpenCL kernel programming language as a library.
 This library is built atop a novel extensibility mechanism that we call \emph{active typechecking and translation} (AT\&T) and detail in Section \ref{att}.
-This mechanism relies on the core idea of representing types as metalanguage objects.
-We refer to such types as \emph{active types} by relation to \emph{active libraries} \cite{activelibraries} (see Section \ref{related}).
-Unlike prior approaches to extensibility that take a language-global view, which hinders interoperability, the AT\&T mechanism guarantees that extensions are composable by construction by limiting their scope to expressions having a type defined by that extension.
+This mechanism relies on the core idea of representing types as first-class objects at compile-time.
+We refer to these objects as \emph{active types} by analogy with \emph{active libraries} \cite{activelibraries} (see Section \ref{related}).
+Unlike prior approaches to extensibility where users globally modify the grammar or semantics of a language, and thus introduce conflicts between extensions, AT\&T guarantees that extensions are composable by construction by pairing new rules with new types and limiting their scope to expressions of that type.
 
@@
 First-class support for each of these has required a new language in the past.
-AT\&T allows compile-time logic to be safely included within libraries, and thus these first-class abstractions can be  used naturally within a single program or workflow.
+AT\&T allows compile-time logic to be safely included within libraries, and thus these first-class abstractions can be coexist naturally within a single program or workflow.
 
@@
 This mode of use, described in Section \ref{compenv}, uses just-in-time specialization to  integrate the widely-adopted \verb|numpy| library (including its internal type system) and OpenCL's host API (by a mechanism also available to other host APIs, such as CUDA's) with the Ace compiler itself to enable the direct invocation of Ace functions from within Python scripts with minimal overhead.
-Achieving this deep level of integration makes use of Ace's representation of types as well.
+Achieving this deep level of integration makes use of Ace's novel type representation as well.
 
@@
 \section{Language Design and Usage}\label{usage}
-\subsection{High-Level Design}
-Static type systems are powerful tools for programming language design and implementation.
-By tracking the type of a value statically, a typechecker can verify the absence of many kinds of errors over all inputs.
-This simplifies and increases the performance of the run-time system, as errors need not be detected dynamically using tag checks and other kinds of assertions.
-Many parallel programming abstractions are defined in terms of, or benefit from, a type system that enforces a communication protocol, ensures the consistency of data and simplifies the dynamics of the run-time system (see Section \ref{usecases} for examples).
-Because \textbf{verifiability} and \textbf{performance} are key criteria and static typing is a core technique, Ace is fundamentally statically-typed.
-
-It is legitimate to ask, however, why dynamically-typed languages are so widely-used in HPC.
-One reason has to do with the criteria of \textbf{ease-of-use}.
-Statically-typed languages, particularly those that HPC programmers are exposed to, can be verbose, requiring explicit type annotations on every function and variable declaration,  separate header files, and other declarations that exist to assist the compiler.
-Cordy identifies the principle of \emph{conciseness} with elimination of
-redundancy and the availability of reasonable defaults \cite{cordy1992hints}.
-The high-level languages used in HPC have, consistent with this principle, avoided most of this overhead.
-Ace was originally conceived to explore the question: does this require run-time mechanisms, or can one express a statically-typed, low-level language at least as powerful as C without the overhead that is typically associated with it?
-
-Rather than designing a new syntax, or modifying the syntax of C, we chose to directly use, without modification, an existing language, Python's, syntax.
-This was not an arbitrary choice but rather a key mechanism by which Ace achieves \textbf{continuity}.
-Python is amongst the most widely-adopted languages in the world, and enjoys similar  ubiquity within HPC specifically \cite{pythonhpc}.
-By directly adopting Python's syntax, Ace's syntax is immediately familiar and acceptable to a significant portion of it's intended audience.
-Moreover, any tools that handle Python source code, including parsers, editors, style checkers and documentation generators, can be used without modification for handling Ace code.
-
-\subsection{Example 1: Hello, World!}
 \begin{codelisting}
@@
 __kernel void main() {
-    printf("Hello, run-time world.");
+    char* hello = "Hello, run-time world!";
+    printf(hello);
 }\end{lstlisting}
@@
 \end{codelisting}
+\subsection{High-Level Design}
+Static type systems are powerful tools for programming language design and implementation.
+By tracking the type of a value statically, a typechecker can verify the absence of many kinds of errors over all inputs.
+This simplifies and increases the performance of the run-time system, as errors need not be detected dynamically using tag checks and other kinds of assertions.
+Many parallel programming abstractions are defined in terms of, or benefit from, a type system that enforces a communication protocol, ensures the consistency of data and simplifies the dynamics of the run-time system (see Section \ref{usecases} for examples).
+Because \textbf{verifiability} and \textbf{performance} are key criteria and static typing is a core technique, Ace is fundamentally statically-typed.
+
+It is legitimate to ask, however, why dynamically-typed languages are so widely-used in HPC.
+Although slow and difficult to reason about, these languages generally excel at satisfying the criteria of \textbf{ease-of-use}.
+More specifically, Cordy identifies the principle of \emph{conciseness} with elimination of
+redundancy and the availability of reasonable defaults \cite{cordy1992hints}.
+Statically-typed languages, particularly those that HPC programmers are exposed to, can be verbose, requiring explicit and often redundant type annotations on each function and variable declaration,  the use of separate header files, explicit template instantiation and other annotations.
+The dynamically-typed languages used in HPC, on the other hand, avoid most of this overhead by relying on support from the run-time system.
+Ace was originally conceived to explore the question: does conciseness require run-time mechanisms, or can one design a statically-typed language with the same low-level semantics of C with syntactic overhead comparable to a high-level scripting language?
+
+Rather than designing a new syntax, or modifying the syntax of C, we chose to directly use, \emph{without modification}, an existing syntax, Python's.
+This was not an arbitrary choice but rather a key decision by which Ace achieves both \textbf{ease-of-use} and \textbf{continuity}.
+Python's whitespace-delimited syntax is widely regarded as both concise and readable, and Python is amongst the most widely-adopted languages in computational science \cite{pythonhpc}.
+By directly adopting Python's syntax, Ace's syntax is immediately \emph{familiar} and \emph{acceptable} to a significant segment of the intended audience.
+Moreover, a key benefit of adopting it without modifications is that any {tools} that handle Python source code, including parsers, editors, style checkers and documentation generators, can be used directly without modification as well.
+
+\subsection{Example: Two-Phase ``Hello, World!''}
+Ace's \textbf{continuity} with Python does not stop at its syntax.
+
+A variant of the standard ``Hello, World!'' example written in Ace is shown in Listing \ref{hello} and its compilation to statically-typed OpenCL code is demonstrated in Listing \ref{helloout}.
 

### 2013-04-26 ace/hello.py

@@ -4,7 +4,8 @@ print "Hello, compile-time world!"
  
 @OpenCL.fn
 def main():
-    printf("Hello, run-time world!")
+	hello = "Hello, run-time world!"
+    printf(hello)
 main = main.compile()
 
 print "Goodbye, compile-time world!"

### 2013-04-26 ace/ace.tex
Revised up through Example 2

@@
 \end{codelisting}
-\subsection{High-Level Design}
+A variant of the standard ``Hello, World!'' example written in Ace is shown in Listing \ref{hello} and its compilation to statically-typed OpenCL kernel code is demonstrated in Listing \ref{helloout}.
+The OpenCL kernel language is a variant of C99 with some additions and restrictions to facilitate execution on GPUs and other accelerators, in addition to conventional CPUs \cite{opencl}.
+We will discuss it further in subsequent examples.
+We emphasize that this module, which we use throughout the paper, is simply a library like any other.
+The core of Ace gives no special treatment to it; it is distributed together with Ace for convenience.
+Aspects of its implementation will be described  in Section \ref{att}.
+
+This example demonstrates several key design decisions that characterize Ace: static typing of run-time behavior, a Python-based syntax, a phase distinction between compile-time and run-time logic, and programmatic compilation.
+We discuss each of these in the next three sections.
+\subsection{Static Typing and Syntax}
 Static type systems are powerful tools for programming language design and implementation.
@@
 Although slow and difficult to reason about, these languages generally excel at satisfying the criteria of \textbf{ease-of-use}.
-More specifically, Cordy identifies the principle of \emph{conciseness} with elimination of
+More specifically, Cordy identified the principle of \emph{conciseness} as elimination of
 redundancy and the availability of reasonable defaults \cite{cordy1992hints}.
-Statically-typed languages, particularly those that HPC programmers are exposed to, can be verbose, requiring explicit and often redundant type annotations on each function and variable declaration,  the use of separate header files, explicit template instantiation and other annotations.
+Statically-typed languages, particularly those that HPC programmers are exposed to, are verbose, requiring explicit and often redundant type annotations on each function and variable declaration, separate header files, explicit template headers and  instantiation and other sorts of annotations.
 The dynamically-typed languages used in HPC, on the other hand, avoid most of this overhead by relying on support from the run-time system.
-Ace was originally conceived to explore the question: does conciseness require run-time mechanisms, or can one design a statically-typed language with the same low-level semantics of C with syntactic overhead comparable to a high-level scripting language?
-
-Rather than designing a new syntax, or modifying the syntax of C, we chose to directly use, \emph{without modification}, an existing syntax, Python's.
-This was not an arbitrary choice but rather a key decision by which Ace achieves both \textbf{ease-of-use} and \textbf{continuity}.
+Ace was first conceived to explore the question: does conciseness require run-time mechanisms, or can one develop a statically-typed language with the same low-level memory and execution model of C but syntactic overhead comparable to a high-level scripting language?
+
+Rather than designing a new syntax, or modifying the syntax of C, we chose to utilize, \emph{without modification}, the syntax of an existing language, Python.
+This choice was not arbitrary, but rather a key means by which Ace achieves both \textbf{ease-of-use} and \textbf{continuity}.
 Python's whitespace-delimited syntax is widely regarded as both concise and readable, and Python is amongst the most widely-adopted languages in computational science \cite{pythonhpc}.
 By directly adopting Python's syntax, Ace's syntax is immediately \emph{familiar} and \emph{acceptable} to a significant segment of the intended audience.
-Moreover, a key benefit of adopting it without modifications is that any {tools} that handle Python source code, including parsers, editors, style checkers and documentation generators, can be used directly without modification as well.
-
-\subsection{Example: Two-Phase ``Hello, World!''}
-Ace's \textbf{continuity} with Python does not stop at its syntax.
-
-A variant of the standard ``Hello, World!'' example written in Ace is shown in Listing \ref{hello} and its compilation to statically-typed OpenCL code is demonstrated in Listing \ref{helloout}.
-
-To demonstrate the somewhat unconventional structure of Ace programs and libraries, we begin in Listing \ref{hello} with a variant of the canonical ``Hello, World!'' example written using the \verb|OpenCL| module.
-We note that this module, which we use throughout the paper in examples, is simply a library like any other.
-The core of Ace gives no special treatment to it; it is distributed with Ace for convenience.
-Aspects of its implementation will be described in Section \ref{att}.
-
-Perhaps the most immediately apparent departure from the standard ``Hello, World!'' example comes on lines 3 and 10 of Listing \ref{hello}, which contain \verb|print| statements that are executed at {\em compile-time}.
-Indeed, Ace programs are Python scripts at the top-level, rather than a list of declarations as in most conventional languages.
-In other words, Python is the {\em compile-time metalanguage} of Ace; Ace programs are embedded within Python.
-A consequence of this choice is that Ace can leverage Python's well-developed package system and  distribution infrastructure directly (Line 1).
-
-The \verb|main| function defined on Lines 5-7 contains a single call to the \verb|printf| primitive, imported from the \verb|ace.OpenCL| module on Line 1.
-The function is introduced using the \verb|@OpenCL.fn| decorator, which indicates that it is a statically-typed Ace function containing run-time logic targeting the \verb|OpenCL| backend, also imported on Line 1.
-Without this decorator, the function would simply be a conventional Python function that could be called only at compile-time (doing so would fail here: \verb|printf| is not a Python primitive.)
-
-On Line 8, this {\em generic function} defined on Lines 5-7, is compiled to produce a {\em concrete function} that we also identify as \verb|main| (overwriting the previous definition).
-The distinction between generic and concrete functions will be explained in our next example.
-For now, we note that only concrete functions are exported when running the Ace compiler, called \verb|acec|, shown operating at the shell in Listing \ref{helloout}.
-The \verb|acec| compiler operates as follows:
-\begin{enumerate}
-\item Executes the provided Python module (\verb|hello.py|)
-\item Produces source code associated with any concrete functions (functions produced using the \verb|compile| method on Line 8 of Listing \ref{hello}) that are in the top-level environment.
-Each backend may produce one or more files (here, \verb|hello.cl| is produced).
-\end{enumerate}
-
-Using the \verb|acec| executable is simply a convenience; the generated code is also available as an attribute of the concrete function immediately after it has been defined, accessed as \verb|main.code|.
-We will show in Section \ref{pyopencl} that, for backend languages with Python bindings, such as OpenCL and C, generic functions can be executed directly, without the intermediate compilation step, if desired.
-
-\subsection{Example 2: Higher-Order Map for OpenCL}
+Moreover, a key benefit of adopting it without modifications is that any {tools} that handle Python source code, including parsers, editors, style checkers and documentation generators, can be used on Ace code without modification.
+
+\subsection{Phase Separation}
 \begin{codelisting}
@@
 \end{codelisting}
-
 \begin{codelisting}
@@
 
-The ``Hello, World!'' example demonstrates the basic structure of Ace programs, but it does not require working with types.
+%To demonstrate the somewhat unconventional structure of Ace programs and libraries, we begin in Listing \ref{hello} with a variant of the canonical ``Hello, World!'' example written using the \verb|OpenCL| module.
+
+Ace's \textbf{continuity} with Python does not stop at its syntax. 
+Perhaps the most immediately apparent departure from the standard ``Hello, World!'' example comes on lines 3 and 10 of Listing \ref{hello}, which contain Python \verb|print| statements that are executed at {\em compile-time}.
+Ace programs and libraries are Python scripts at the top-level, rather than a list of declarations as in most conventional languages.
+In other words, Python is the {\em compile-time metalanguage} of Ace.
+
+A consequence of this choice is that Ace can leverage Python's well-developed package system and associated distribution infrastructure directly (e.g.
+Line 1).
+This serves to address another key \textbf{ease-of-use} issue often associated with C-based languages: the fragility of the preprocessor-based packaging system they historically have relied upon.
+
+Ace functions defining run-time logic are introduced by a decorator.
+The \verb|@OpenCL.fn| decorator on Line 5 indicates that the \verb|main| function is a statically-typed Ace function targeting the OpenCL backend for code generation (Section \ref{backend}).
+Without this decorator, the function would simply be a conventional Python function that could be called only at compile-time.
+Doing so would fail here: \verb|printf| is not a Python function.
+Rather, it is a compile-time Python object representing an OpenCL primitive residing in the \verb|OpenCL| module.
+It has an associated Ace type that controls what types of input it can receive and the type of its output given its input type (see Section \ref{att}).
+
+\subsection{Programmatic Compilation}
+As defined on Lines 5-7, \verb|main| is a \emph{generic function}.
+This terminology in Ace is used to indicate a function for which types have not yet been assigned to arguments.
+Invoking the \verb|compile| method on a generic function with a sequence of argument types invokes the active typechecking and translation mechanism we will describe in Section 3 to produce a \emph{concrete function} -- one with a single type for each argument, internal variable (such as \verb|hello|) and the return value.
+In Listing \ref{hello}, we name this concrete function \verb|main|, overwriting the generic function of the same name that it is derived from.
+The compiler does nothing apparently interesting: there are no arguments, the single internal variable takes the OpenCL string type \verb|char*| from its value and the return type is \verb|void|.
+
+Before moving on to more interesting examples, let us discuss the \verb|acec| compiler shown operating at the shell in Listing \ref{helloout}. The \verb|acec| compiler operates in two steps:
+\begin{enumerate}
+\item Executes the provided Python file (\verb|hello.py|) which contains Ace functions and (in future examples) types and compile-time code generation logic.
+\item Produces source code for concrete functions (produced using the \verb|compile| method) in the top-level Python environment and any other concrete functions, type declarations and other program items required by or generated by these functions.
+This may produce one or more files (here, just \verb|hello.cl|).
+\end{enumerate}
+
+We will show in Section \ref{pyopencl} that for backends with Python bindings, such as OpenCL, CUDA and C, generic functions can be executed directly, without this explicit compilation step to concrete functions, if desired.
+
+\subsection{Example 2: Higher-Order Map}
+The ``Hello, World!'' example demonstrates the structure of Ace programs, but it does not require working with types.
 Listing \ref{map} shows an imperative, data-parallel map primitive written using the OpenCL library introduced above.
-In OpenCL, users can define functions, called {\em kernels}, that execute across thousands of threads.
+To review, in OpenCL users can define functions, called {\em kernels}, that execute across thousands of threads.
 Each kernel has access to a unique index, called its {\em global id}, which can be used by the programmer to ensure that each thread operates on different parts of the input data (Line 5).
@@
 
-As above, \verb|map| is a {\em generic function} (specifically, an instance of the class \verb|ace.GenericFn|).
+As above, \verb|map| is a {\em generic function} (specifically, it is an instance of the class \verb|ace.GenericFn|).
 This means that it's arguments have not been assigned types.

### 2013-04-26 ace/ace.tex
Revised up through type propagation

@@
 \lstinputlisting[style=OpenCL]{listing5.cl}
-\caption{[\texttt{listing\ref{mapout}.cl}] The OpenCL code generated by running \texttt{acec listing2.py}.}
+\caption{[\texttt{listing\ref{mapadd5dbl}.cl}] The OpenCL code generated by running \texttt{acec listing\ref{mapadd5dbl}.py}.}
 \label{mapout}
@@
 To review, in OpenCL users can define functions, called {\em kernels}, that execute across thousands of threads.
-Each kernel has access to a unique index, called its {\em global id}, which can be used by the programmer to ensure that each thread operates on different parts of the input data (Line 5).
+Each kernel has access to a unique index, called its {\em global id}, which can be to ensure that each thread operates on different parts of the input data (Line 5).
 The \verb|map| kernel defined in Listing \ref{map} applies a transfer function, \verb|f|, to the element of the input array, \verb|input|, corresponding to its global id.
@@
 As above, \verb|map| is a {\em generic function} (specifically, it is an instance of the class \verb|ace.GenericFn|).
-This means that it's arguments have not been assigned types.
-Indeed, the functionality given by the \verb|map| definition is applicable to many combinations of types for \verb|input| and \verb|output| and functions, \verb|f|.
-In this sense, \verb|map| is actually a family of functions defined for all types assignments for \verb|input|, \verb|output| and \verb|f| for which the operations in the function's body are well-defined.
-
-Running \verb|acec listing3.py| would produce no output, however.
-To create a {\em concrete function} (an instance of the class \verb|ace.ConcreteFn|) that can be emitted by the compiler, types must be assigned to each of the arguments of any externally callable functions.
-Listing \ref{mapadd5dbl} shows how to use the \verb|compile| method to specialize \verb|map| in two different ways to apply the \verb|add5| function, defined on Lines 4-6, to arrays  that reside in global memory (OpenCL has a notion of four different memory spaces).
-Line 9 produces a version specialized for arrays of \verb|double|s and Line 10 produces a version for arrays of \verb|int|s.
-The output of compilation is shown in Figure \ref{mapout}.
+This means that its arguments have not been assigned types.
+The functionality given by the \verb|map| definition is in fact applicable to many combinations of types for \verb|input| and \verb|output| and functions, \verb|f|.
+In this sense, \verb|map| is actually a \emph{family} of functions defined for all types assignments for \verb|input|, \verb|output| and \verb|f| such that the operations in the function's body are well-defined.
+
+Running \verb|acec listing3.py| would produce no output.
+To create a {\em concrete function} (that is, an instance of the class \verb|ace.ConcreteFn|) that can be emitted by the compiler, types must be assigned to each of the arguments.
+Listing \ref{mapadd5dbl} shows how to use the \verb|compile| method to specialize \verb|map| in \emph{two} different ways to apply the \verb|add5| function, defined on Lines 4-6, to arrays  that reside in global memory (OpenCL associates a memory space with pointer types).
+Line 9 results in a version specialized for arrays of \verb|double|s and Line 10 results in a version for arrays of \verb|int|s.
+The output of compilation is shown in Listing \ref{mapout}.
 
 \subsection{Types as Metalanguage Objects}
-The types of the arguments of the functions being compiled on Lines 4.9 and 4.10, if written directly in OpenCL as in Listing \ref{mapout}, are \verb|__global double*| and \verb|__global int*|, respectively.
-The corresponding types in Ace are \verb|global_ptr(double)| and \verb|global_ptr(int)|, abbreviated for convenience as \verb|A| and \verb|B| respectively on Line 8 (note that here, standard variable assignment in the metalanguage is serving the role that a \verb|typedef| declaration would fill in a C-like language.)
-
-\textbf{Types in Ace are values in the metalanguage}, Python.
-More specifically, types are {\em object instances} of user-defined classes that inherit from the Ace-provided \verb|ace.Type| class.
-For example, \verb|global_ptr(double)| is an instance of \verb|ace.OpenCL.GlobalPtrType| instantiated with the target type, \verb|double|, as a constructor argument.
-The types \verb|double| and \verb|int| (imported from \verb|ace.OpenCL| on Line 4.2) are instances of \verb|ace.OpenCL.FloatType| and \verb|ace.OpenCL.IntegerType|, respectively.
-As we will describe further in Section \ref{att}, this notion of types as metalanguage objects is a key to Ace's flexibility.
-
-\subsection{Type Propagation and Higher-Order Functions}
-The type assigned to the third argument, \verb|f|, on Lines 4.9 and 4.10, is given as \verb|add5.ace_type|.
-The \verb|ace_type| attribute of a generic function is an instance of \verb|ace.GenericFnType|, the type of Ace generic functions, that corresponds to that particular function, \verb|add5| in this case.
-Generic functions are compiled automatically when they are called from another function.
-That is, when the compiler encounters the call to \verb|f| inside \verb|map| when compiling \verb|map_add5_double|, it compiles a version of \verb|add5| specialized to the \verb|double| type (seen on Line 5.3), and similarly with the \verb|int| type when compiling \verb|map_add5_int| (on Line 5.16, automatically given a unique name to avoid conflicts).
+The \verb|compile| method assigns types to the arguments of a generic function.
+In Listing \ref{mapadd5dbl}, the types we are using are given shoter names, for convenience, on Line 8 (that is, variables in the metalanguage can be used like \verb|typedef| is used in a C-like language).
+The types \verb|int| and \verb|double| imported from the \verb|OpenCL| module correspond to the OpenCL types of the same name.
+The types \verb|gptr(int)| and \verb|gptr(double)| correspond to \verb|__global int*| and \verb|__global double*|.
+That is, \verb|gptr| can be understood as a type-indexed family of types.
+
+These types are \emph{objects in the metalanguage}, Python.
+More specifically, types are {instances} of user-defined classes that inherit from the Ace-provided \verb|ace.Type| class.
+For example, \verb|gptr(double)| is an instance of \verb|OpenCL.GlobalPtrType| instantiated with the target type, \verb|double|, as a constructor argument.
+The types \verb|double| and \verb|int| are instances of \verb|OpenCL.FloatType| and \verb|OpenCL.IntegerType|, respectively.
+This notion of types as metalanguage objects is key to the Ace compilation model and also enables other mechanisms that we will discuss in subsequent sections.
+
+\subsection{Type Propagation}
+\begin{codelisting}
+\lstinputlisting{listing6.py}
+\caption{\texttt{[listing6.py]} A function demonstrating whole-function type inference when multiple values with differing types are assigned to a single variable.}
+\label{inference}
+\end{codelisting}
+The type assigned to the third argument, \verb|f|, on both Lines 4.9 and 4.10, is \verb|add5.ace_type|.
+The \verb|ace_type| attribute of a generic function is an instance of \verb|ace.GenericFnType|, the type of Ace generic functions.
+Ace generic functions are compiled to concrete functions automatically at all internal call sites.
+That is, when the compiler encounters the call to \verb|f| inside \verb|map| when compiling \verb|map_add5_double|, it compiles a version of \verb|add5| specialized to the \verb|double| type (seen on Line 5.3), and similarly when compiling \verb|map_add5_int| (on Line 5.16, automatically given a unique name to avoid conflicts).
 This mechanism is called {\em type propagation}.
-In other words, we did not need to use \verb|add5.compile(double)| before compiling \verb|map_add5_dbl| (although we could if we would like).
-Only functions that are never called in the process of compiling other functions in a module need type information explicitly provided, whereas \verb|add5| is a function that is only called within \verb|map| in Listing \ref{mapadd5dbl}.
-
-We note that this scheme allows for a form of higher-order functional programming even when targeting languages, like OpenCL, that have no support for higher-order functions (OpenCL, unlike C99, does not even support function pointers).
-This works because the \verb|ace.GenericFnType| for one function, such as \verb|add5|, is not the same as the \verb|ace.GenericFnType| for a superficially similar function, such as \verb|add6| (defined as one would expect).
-To put it in type theoretic terms, \verb|ace.GenericFnType|s are singleton types, {\em uniquely inhabited} by a single generic function, and thus it is impractical to use them as full first-class values (i.e.
-they cannot put into a run-time array.) In practice, this is rarely a problem, particularly in parallel programming where explicit specialization is already common in order to avoid the potential performance penalty associated with a function pointer dereference per call.
-We note that fully-featured first-class functions can be implemented when targeting a backend that supports them, such as C99, or simulated by using a jump table implementation in OpenCL, but we do not discuss this further due to lack of space.
+We did not need to use \verb|add5.compile(double)| before compiling \verb|map_add5_dbl| because only functions that are never called in the process of compiling other functions in a module need type information explicitly provided, supporting \textbf{ease-of-use} by increasing conciseness.
+
+In effect, this scheme allows for a form of higher-order functional programming even when targeting languages, like OpenCL, that have no support for higher-order functions (OpenCL, unlike C99, does not support function pointers).
+This works because the \verb|ace.GenericFnType| for one function, such as \verb|add5|, is not equal to  the \verb|ace.GenericFnType| for a superficially similar function, such as \verb|add6| (defined as one would expect).
+To put it in type theoretic terms, \verb|ace.GenericFnType|s are singleton types, {\em uniquely inhabited} by a single generic function.
+A consequence of this is that they cannot be used as first-class values (i.e.
+they cannot written into an array).
+This is often valuable, particularly in parallel programming where compile-time specialization is valuable to avoid the \textbf{performance} and \textbf{ease-of-use} issues that occur if using function pointers.
+Concrete functions, on the other hand, can be given a true function type (e.g.
+\verb|add5| could be compiled to a concrete function with type \verb|int| $\rightarrow$ \verb|int|) if targeting a backend that supports them, such as C99, or by using an integer-indexed jump table in OpenCL (we have not implemented this mechanism using Ace as of the time of writing, but do not anticipate difficulties).
+
+Type propagation via generic functions can be compared to template specialization in C++, where both the necessary template headers (containing nested template parameters to support function passing) and template specializations are inferred automatically from usage.
+This significantly simplifies a sophisticated feature of C++ and introduces it to OpenCL and C, which do not support templates.
+% Is this really any more concise than using C + templates? Etc. relation to types. Conciseness.
 
@@
 Note that this example would also work correctly if the assignments to \verb|y| were replaced with \verb|return| statements (in other words, the return value of a function is treated as an assignable for the purpose of type inference).
-
-\begin{codelisting}
-\lstinputlisting{listing6.py}
-\caption{\texttt{[listing6.py]} A function demonstrating whole-function type inference when multiple values with differing types are assigned to a single variable.}
-\label{inference}
-\end{codelisting}
 

### 2013-04-26 ace/listing4.py

@@ -1,10 +1,10 @@
 from listing3 import map
-from ace.OpenCL import global_ptr, double, int
+from ace.OpenCL import gptr, double, int
 
 @OpenCL.fn
 def add5(x):
     return x + 5
 
-A = global_ptr(double); B = global_ptr(int)
-map_add5_dbl = map.compile(A, A, add5.ace_type)
-map_add5_int = map.compile(B, B, add5.ace_type)
+D = gptr(double); I = gptr(int); A = add5.ace_type
+map_add5_dbl = map.compile(D, D, A)
+map_add5_int = map.compile(I, I, A)

### 2013-04-26 ace/ace.tex
Section 2 done

@@
 The types \verb|gptr(int)| and \verb|gptr(double)| correspond to \verb|__global int*| and \verb|__global double*|.
-That is, \verb|gptr| can be understood as a type-indexed family of types.
 
@@
 More specifically, types are {instances} of user-defined classes that inherit from the Ace-provided \verb|ace.Type| class.
-For example, \verb|gptr(double)| is an instance of \verb|OpenCL.GlobalPtrType| instantiated with the target type, \verb|double|, as a constructor argument.
+For example, \verb|gptr(double)| is an instance of \verb|OpenCL.PtrType| instantiated with the target type, \verb|double|, and memory space, \verb|__global|, as constructor arguments.
 The types \verb|double| and \verb|int| are instances of \verb|OpenCL.FloatType| and \verb|OpenCL.IntegerType|, respectively.
@@
 \subsection{Type Propagation}
-\begin{codelisting}
-\lstinputlisting{listing6.py}
-\caption{\texttt{[listing6.py]} A function demonstrating whole-function type inference when multiple values with differing types are assigned to a single variable.}
-\label{inference}
-\end{codelisting}
 The type assigned to the third argument, \verb|f|, on both Lines 4.9 and 4.10, is \verb|add5.ace_type|.
@@
 To put it in type theoretic terms, \verb|ace.GenericFnType|s are singleton types, {\em uniquely inhabited} by a single generic function.
-A consequence of this is that they cannot be used as first-class values (i.e.
-they cannot written into an array).
-This is often valuable, particularly in parallel programming where compile-time specialization is valuable to avoid the \textbf{performance} and \textbf{ease-of-use} issues that occur if using function pointers.
+A consequence of this is that they are not useful as first-class values (i.e.
+they cannot be written into a collection).
+This is often valuable, particularly in parallel programming where compile-time specialization is valuable to avoid \textbf{performance} and \textbf{ease-of-use} issues that occur when using function pointers.
+
 Concrete functions, on the other hand, can be given a true function type (e.g.
-\verb|add5| could be compiled to a concrete function with type \verb|int| $\rightarrow$ \verb|int|) if targeting a backend that supports them, such as C99, or by using an integer-indexed jump table in OpenCL (we have not implemented this mechanism using Ace as of the time of writing, but do not anticipate difficulties).
-
-Type propagation via generic functions can be compared to template specialization in C++, where both the necessary template headers (containing nested template parameters to support function passing) and template specializations are inferred automatically from usage.
+\verb|add5| could be compiled to a concrete function with type \verb|int| $\rightarrow$ \verb|int|) if targeting a backend that supports them, such as C99, or by using an integer-indexed jump table in OpenCL (we have not implemented this mechanism using Ace as of the time of writing, but do not anticipate difficulties in doing so).
+
+Type propagation via generic functions can be compared to template specialization in C++, where both the template headers (containing nested template parameters for function arguments) and specialization parameters at any call sites are inferred automatically from usage.
 This significantly simplifies a sophisticated feature of C++ and introduces it to OpenCL and C, which do not support templates.
+Other uses for C++ templates are subsumed by the metaprogramming features discussed in Section \ref{metaprogramming}.
 % Is this really any more concise than using C + templates? Etc. relation to types. Conciseness.
 
-\subsection{Type Inference}
-On Line 5 in the generic \verb|map| function in Listing \ref{map}, the variable \verb|gid| is initialized with the result of calling the \verb|get_global_id| primitive (the argument, \verb|0|, is not important for our purposes.)  Note that the type for the \verb|gid| variable is never listed explicitly.
-This is because Ace supports a form of {\em whole-function type inference}.
-In this case, \verb|gid| will be assigned type \verb|size_t| because that is the return type of \verb|get_global_id| (as defined in the OpenCL specification \cite{opencl}, which the \verb|ace.OpenCL| module follows.) The result can be observed on Lines 11 and 24 in Listing \ref{mapout}.
-
-Inference is not restricted within single assignments, as in the \verb|map| example.
-Multiple assignments to the same variable with values of differing types, or multiple return statements, can be unified such that the variable or return type is given a common supertype.
-For example, in the \verb|threshold_scale| function defined in Listing \ref{inference}, the variable \verb|y| in the first branch of the conditional is assigned the \verb|int| literal \verb|0|.
-However, in the second branch of the loop, its type depends on the types of \verb|x| and \verb|scale|.
-We show two choices for these types on Lines 10 and 11.
-However, type inference correctly unifies these two types according to OpenCL's C99-derived rules governing numeric types (which are defined by the user in the \verb|OpenCL| module, as we will describe in Section \ref{att}).
-We test this programmatically on Lines 12 and 13.
+\subsection{Whole-Function Type Inference}
+\begin{codelisting}
+\lstinputlisting{listing6.py}
+\caption{\texttt{[listing6.py]} A function demonstrating whole-function type inference when multiple values with differing types are assigned to a single identifier, \texttt{y}.}
+\label{inference}
+\end{codelisting}
+
+On Line 5 in the generic \verb|map| function in Listing \ref{map}, the variable \verb|gid| is initialized with the result of calling the OpenCL primitive \verb|get_global_id|.
+The type for \verb|gid| is never given explicitly.
+This is a simple case of Ace's more general {\em whole-function type inference}.
+In this case, \verb|gid| will be inferred to have type \verb|size_t| because that is the return type of \verb|get_global_id| (as defined in the OpenCL specification \cite{opencl}, which the \verb|ace.OpenCL| module follows).
+The result can be observed on Lines 11 and 24 in Listing \ref{mapout}.
+
+Inference is not restricted within single assignments, as in the \verb|map| example, however.
+Multiple assignments to the same identifier with values of differing types, or multiple return statements, can be combined if the types in each case are compatible with one another (e.g.
+by a subtyping relation or an implicit coercion).
+In Listing \ref{inference}, the \verb|threshold_scale| function assigns different values to \verb|y| in each branch of the conditional.
+In the first branch, the value \verb|0| is an \verb|int| literal.
+However, in the second branch of the loop, the type depends on the types of both arguments, \verb|x| and \verb|scale|.
+We show two choices for these types on Lines 11 and 12.
+Type inference correctly combines these two types according to OpenCL's C99-derived rules governing numeric types (defined by the user in the \verb|OpenCL| module, as we will describe in Section \ref{att}).
+We can verify this programmatically on Lines 12 and 13.
 Note that this example would also work correctly if the assignments to \verb|y| were replaced with \verb|return| statements (in other words, the return value of a function is treated as an assignable for the purpose of type inference).
-
 \subsection{Annotation and Extension Inference}
@@
 Users can annotate functions that meet certain requirements to be callable from the host with the \verb|__kernel| attribute.
-The \verb|Ace.OpenCL.OpenCL| backend is able to check these requirements and add this annotation automatically.
-Several types (notably, \verb|double|) and specialized functions require that an OpenCL extension be enabled with a \verb|#pragma| when used.
+The \verb|OpenCL| backend is able to check these requirements and add this annotation automatically.
+Several types (notably, \verb|double|) and specialized functions require that an OpenCL extension be enabled with a \verb|#pragma| if used.
 The OpenCL backend automatically detects many of these cases as well and adds the appropriate \verb|#pragma| declaration.
 An example of this feature can be seen on Line 5.1, where the use of the \verb|double| type triggers the insertion of an appropriate \verb|#pragma| automatically.
-Ace is designed to allow backends to observe the results of the type checking process to support this form of inference.
-
-\subsection{Metaprogramming in Ace}
-\begin{codelisting}
-\lstinputlisting[commentstyle=\color{mauve}]{listing7.py}
-\caption{[\texttt{listing7.py}] Metaprogramming with Ace, showing how to construct generic functions from both strings and abstract syntax trees, and how to manipulate syntax trees at compile-time.}
-\label{metaprogramming}
-\end{codelisting}
-Metaprogramming refers to the practice of writing programs that manipulate other programs.
-There are a number of use cases for this technique, including domain-specific optimizations and code generation for programs with a repetitive structure that cannot easily be captured using available abstractions \cite{pyopencl}.
-OpenCL in particular relies on code generation as a fundamental mechanism, which is cited as justification for its lack of support for higher-order programming.
-Ace supports programmatic compilation and higher-order constructs, as described above, and a flexible language extension mechanism, which we describe below, so several use cases for metaprogramming have been eliminated.
-However, cases where this form of metaprogramming include programmatic function specialization (Listing \ref{metaprogramming}) and modular simulation orchestration, described in Section \ref{clegans}.
-
-On Lines 4-7 of Listing \ref{metaprogramming}, an Ace function is constructed from a string containing its source using the \verb|from_source| variant of the \verb|OpenCL.fn| method.
-It can also be constructed directly from an abstract syntax tree (AST), as implemented by the Python standard \verb|ast| package, using the \verb|from_ast| variant of \verb|fn|, demonstrated on Line 10.
-The AST here is generated programmatically by calling the \verb|specialize| function, which produces a copy of the syntax tree of the \verb|plus| function with the argument, \verb|b|, eliminated and its uses replaced with a constant, \verb|5|.
-This transformation as well as some others are distributed in \verb|ace.astx| for convenience.
-\subsection{Direct Invocation from Python}\label{direct}
-
-\begin{codelisting}
-\lstinputlisting{listing8.py}
-\caption{[\texttt{listing\ref{py}.py}] A full OpenCL program using the \texttt{Ace.OpenCL} Python bindings, including data transfer to and from a device and direct invocation of a generic function, \texttt{map}, as a kernel without explicit compilation.}
-\label{py}
-\end{codelisting}
-
-As discussed in the Introduction, a common workflow for professional end-users involves the use of a high-level scripting language for overall workflow orchestration and small-scale data analysis and visualization, paired with a low-level language for performance-critical sections.
-Python is already widely used by professional end-users as a high-level scripting language and also features mature support for calling into code written in low-level languages.
-Developers can call into native libraries using its foreign function interface (FFI),  or by using a wrapper library like \verb|pycuda| for code compiled with CUDA, a proprietary language similar to OpenCL  specifically targeting nVidia GPU hardware.
-Although CUDA's compilers are separate executables on the system, the OpenCL language was designed for this workflow, in that it exposes the compiler directly as an API.
-The \verb|pyopencl| module exposes this API as well as the OpenCL memory management API to Python.
-With both \verb|pyopencl| and \verb|pycuda|, developers generate source code as strings, compile it programmatically, then execute it using the run-time APIs that each library provides \cite{pyopencl}.
-
-Ace supports this workflow as an alternative to using the \verb|acec| compiler as described above to generate  source code directly from the shell.
-For the OpenCL backend, these bindings are exposed as a wrapper on top of \verb|pyopencl| called \verb|Ace.OpenCL.bindings|.
-Both generic functions and concrete functions written for a backend that supports the direct execution interface (thus far, \verb|OpenCL| and \verb|C99|) can be called like regular Python functions.
-An example of this for the generic \verb|map| function defined in Listing \ref{map} is shown in Listing \ref{py}, with the call itself on Lines 9-10.
-The first two arguments to \verb|map| are OpenCL buffers, generated using a simplified wrapper to the \verb|pyopencl| APIs on Lines 7-8.
-This wrapper associates type information with each buffer, similarly to \verb|numpy|, and this is used to implicitly compile \verb|map| as appropriate the first time it is called for any given combination of input types.
-Explicit calls to the \verb|compile| method, as we have been showing thusfar, are unnecessary if using this method of invocation.
-The final two keyword arguments on Line 10 are parameters that OpenCL requires for execution that determine the number of threads (called the {\em global size}) and thread grouping (the {\em local size}).
-
-By way of comparison, the same program written using the OpenCL C API directly is an order of magnitude larger and correspondingly more complex.
-A full implementation of the logic of \verb|map| written using the \verb|pyopencl| bindings and metaprogramming techniques as described in \cite{pyopencl} is twice as large and significantly more difficult to comprehend than the code we have shown thus far.
-Not shown are several additional conveniences, such as delegated kernel sizing and \verb|In| and \verb|Out| constructs that can reduce the size and improve the clarity of this code further; due to a lack of space, the reader is referred to the language documentation for additional details on these features.
 
 \section{Active Typechecking and\\Translation (AT\&T)}\label{att}
-\subsection{Active Translation}\label{translation}
-\subsection{Use Cases}\label{usecases}
-\section{Ace: A Compilation Environment}\label{compenv}
-\section{Case Study}\label{casestudy}
-\section{Related Work}\label{related}
-\section{Discussion}\label{discussion}
-%We argue that many of the barriers to adoption faced by new abstractions stem from the fact that new abstractions are too-often developed and distributed together with a new programming language.
-Some prominent cases where such a {\em language-oriented approach} \cite{languageoriented} has been followed include:
-%\begin{itemize}
-%\item UPC, CAF, X10 and Chapel, built around first-class support for a partitioned global address space (PGAS) programming model; 
-%\item Go, Erlang, Charm++, Occam and Axum, built around first-class support for message passing; and
-%\item CUDA, OpenCL, Cilk and NESL, built around first-class support for data parallelism.
-%\end{itemize}
-%
-%The benefits of taking this approach are clear: first-class language, compiler and run-time support for a programming abstraction can improve performance and enable a more natural programming style.
-Empirical evidence also supports this notion that developers prefer first-class support for a parallel abstraction over a library-based solution, if other factors are held constant \cite{ppfirstclass}.
-%
-%Unfortunately, adopting a specialized research language comes with a host of ancillary challenges  \cite{adoptionTR,socioplt}.
-These are significant enough that programming abstractions that are distributed as libraries for a well-established language, such as MapReduce, Global Arrays and MPI, have prevailed in practice, despite the limited expressivity of library-based abstractions.
-This is balanced by several factors: developers can progressively introduce parallelism into existing programs, leverage a familiar and well-developed general-purpose programming language when not directly invoking parallel primitives, and utilize arbitrary combinations of  abstractions within a single program without needing to navigate the friction inherent to language boundaries.
-%
-%Ace aims to bridge the language-oriented and library-oriented approaches by allowing developers to define and distribute a broad range of new {\em first-class} types and their operations directly within user libraries, using a mechanism called {\em active typechecking and translation (\ATT)}.
-Based on the notion of active libraries \cite{activelibraries}, \ATT~is designed to allow developers to  extend the semantics, but not the syntax, of a language.
-The Ace grammar is a nearly-complete subset of Python's  and we show that it is flexible enough to support all of the primitives of the C99 and OpenCL languages in a natural manner, when aided by a form of type inference that we describe.
-We further supplement these primitives with higher-level user-defined primitives, examples of which we discuss in Section \ref{abstractions}.
-Notably, all of these primitives are imported via the normal library loading mechanism; they are not defined within domain-specific languages, nor as extensions for a particular compiler implementation.
-%
-%Although the \ATT~mechanism, a key feature of Ace, is designed primarily for use by researchers and domain experts, Ace is intended to be highly practical and ready for adoption by professional end-user developers today.
-It is based on the Python programming language, which enjoys broad adoption in this community.
-Indeed, the language itself is implemented entirely as a Python library, and many existing tools, such as editor modes, documentation tools and style checkers, can be used without modification.
-%Ace functions are generated programmatically and can be called directly from Python, with numeric arrays as arguments, so developers can integrate Ace immediately within their existing workflows for portions of code that may benefit from static compilation and support for parallelism.
-%We demonstrate the practicality of Ace in Section \ref{clegans} by describing an Ace-based framework that supports specifying and orchestrating families of large-scale neurobiological circuit simulations on clusters of GPUs.
-
-\section{Structure and Usage}
-%Ace is an {\em actively-typed, language-integrated compilation environment}. Let us begin by defining each of these terms. 
-%A {\em compilation environment} is a system that allows developers to programmatically control aspects of compilation directly.
-The language used to do so is called the {\em metalanguage} while the grammar that programs themselves are written in is called the {\em source grammar}.
-The metalanguage of Ace is Python and the source grammar is a significant subset of Python's own grammar.
-Programs and metalanguage code can be interleaved within the same file, so Ace is {\em language-integrated}.
-Ace programs are statically-typed, and types themselves are objects in the metalanguage.
-They can be equipped by developers with methods that control how operators associated with them are typechecked and compiled.
-Because of the relationship between this mechanism and the idea of {\em active libraries} \cite{activelibraries}, discussed further in Section \ref{related}, we call them {\em active types}, and thus Ace is {\em actively-typed}.
-\section{Ace for Researchers}
 Thus far, we have been largely discussing the OpenCL module in our examples.
@@
 \subsection{Active Typechecking and Translation (\ATT)}
+\begin{codelisting}
+\lstinputlisting{listing9.py}
+\caption{\texttt{[ace.OpenCL]} A portion of the implementation of OpenCL pointer types implementing subscripting logic using the Ace extension mechanism, AT\&T.}
+\label{pointers}
+\end{codelisting}
 We now explain how new primitive types and operators can be defined in Ace.
@@
 \end{itemize}
-\begin{codelisting}
-\lstinputlisting{listing9.py}
-\caption{\texttt{[listing9.py]} A portion of the implementation of OpenCL pointer types implementing subscripting logic using the Ace extension mechanism.}
-\label{pointers}
-\end{codelisting}
 
@@
 
-The relevant portion of \verb|ace.OpenCL.GlobalPtrType| is shown in Listing \ref{pointers}.
+The relevant portion of \verb|OpenCL.GlobalPtrType| is shown in Listing \ref{pointers}.
 The \verb|verify_Subscript| method on Line 8 receives a context and the syntax tree of the node itself as input.
@@
 
+\subsection{Active Translation}\label{translation}
+\subsection{Use Cases}\label{usecases}
+\section{Ace: A Compilation Environment}\label{compenv}
+\subsection{Metaprogramming in Ace}
+\begin{codelisting}
+\lstinputlisting[commentstyle=\color{mauve}]{listing7.py}
+\caption{[\texttt{listing7.py}] Metaprogramming with Ace, showing how to construct generic functions from both strings and abstract syntax trees, and how to manipulate syntax trees at compile-time.}
+\label{metaprogramming}
+\end{codelisting}
+Metaprogramming refers to the practice of writing programs that manipulate other programs.
+There are a number of use cases for this technique, including domain-specific optimizations and code generation for programs with a repetitive structure that cannot easily be captured using available abstractions \cite{pyopencl}.
+OpenCL in particular relies on code generation as a fundamental mechanism, which is cited as justification for its lack of support for higher-order programming.
+Ace supports programmatic compilation and higher-order constructs, as described above, and a flexible language extension mechanism, which we describe below, so several use cases for metaprogramming have been eliminated.
+However, cases where this form of metaprogramming include programmatic function specialization (Listing \ref{metaprogramming}) and modular simulation orchestration, described in Section \ref{clegans}.
+
+On Lines 4-7 of Listing \ref{metaprogramming}, an Ace function is constructed from a string containing its source using the \verb|from_source| variant of the \verb|OpenCL.fn| method.
+It can also be constructed directly from an abstract syntax tree (AST), as implemented by the Python standard \verb|ast| package, using the \verb|from_ast| variant of \verb|fn|, demonstrated on Line 10.
+The AST here is generated programmatically by calling the \verb|specialize| function, which produces a copy of the syntax tree of the \verb|plus| function with the argument, \verb|b|, eliminated and its uses replaced with a constant, \verb|5|.
+This transformation as well as some others are distributed in \verb|ace.astx| for convenience.
+\subsection{Direct Invocation from Python}\label{direct}
+
+\begin{codelisting}
+\lstinputlisting{listing8.py}
+\caption{[\texttt{listing\ref{py}.py}] A full OpenCL program using the \texttt{Ace.OpenCL} Python bindings, including data transfer to and from a device and direct invocation of a generic function, \texttt{map}, as a kernel without explicit compilation.}
+\label{py}
+\end{codelisting}
+
+As discussed in the Introduction, a common workflow for professional end-users involves the use of a high-level scripting language for overall workflow orchestration and small-scale data analysis and visualization, paired with a low-level language for performance-critical sections.
+Python is already widely used by professional end-users as a high-level scripting language and also features mature support for calling into code written in low-level languages.
+Developers can call into native libraries using its foreign function interface (FFI),  or by using a wrapper library like \verb|pycuda| for code compiled with CUDA, a proprietary language similar to OpenCL  specifically targeting nVidia GPU hardware.
+Although CUDA's compilers are separate executables on the system, the OpenCL language was designed for this workflow, in that it exposes the compiler directly as an API.
+The \verb|pyopencl| module exposes this API as well as the OpenCL memory management API to Python.
+With both \verb|pyopencl| and \verb|pycuda|, developers generate source code as strings, compile it programmatically, then execute it using the run-time APIs that each library provides \cite{pyopencl}.
+
+Ace supports this workflow as an alternative to using the \verb|acec| compiler as described above to generate  source code directly from the shell.
+For the OpenCL backend, these bindings are exposed as a wrapper on top of \verb|pyopencl| called \verb|Ace.OpenCL.bindings|.
+Both generic functions and concrete functions written for a backend that supports the direct execution interface (thus far, \verb|OpenCL| and \verb|C99|) can be called like regular Python functions.
+An example of this for the generic \verb|map| function defined in Listing \ref{map} is shown in Listing \ref{py}, with the call itself on Lines 9-10.
+The first two arguments to \verb|map| are OpenCL buffers, generated using a simplified wrapper to the \verb|pyopencl| APIs on Lines 7-8.
+This wrapper associates type information with each buffer, similarly to \verb|numpy|, and this is used to implicitly compile \verb|map| as appropriate the first time it is called for any given combination of input types.
+Explicit calls to the \verb|compile| method, as we have been showing thusfar, are unnecessary if using this method of invocation.
+The final two keyword arguments on Line 10 are parameters that OpenCL requires for execution that determine the number of threads (called the {\em global size}) and thread grouping (the {\em local size}).
+
+By way of comparison, the same program written using the OpenCL C API directly is an order of magnitude larger and correspondingly more complex.
+A full implementation of the logic of \verb|map| written using the \verb|pyopencl| bindings and metaprogramming techniques as described in \cite{pyopencl} is twice as large and significantly more difficult to comprehend than the code we have shown thus far.
+Not shown are several additional conveniences, such as delegated kernel sizing and \verb|In| and \verb|Out| constructs that can reduce the size and improve the clarity of this code further; due to a lack of space, the reader is referred to the language documentation for additional details on these features.
+
+\section{Case Study}\label{casestudy}
+\section{Related Work}\label{related}
+\section{Discussion}\label{discussion}
+%We argue that many of the barriers to adoption faced by new abstractions stem from the fact that new abstractions are too-often developed and distributed together with a new programming language.
+Some prominent cases where such a {\em language-oriented approach} \cite{languageoriented} has been followed include:
+%\begin{itemize}
+%\item UPC, CAF, X10 and Chapel, built around first-class support for a partitioned global address space (PGAS) programming model; 
+%\item Go, Erlang, Charm++, Occam and Axum, built around first-class support for message passing; and
+%\item CUDA, OpenCL, Cilk and NESL, built around first-class support for data parallelism.
+%\end{itemize}
+%
+%The benefits of taking this approach are clear: first-class language, compiler and run-time support for a programming abstraction can improve performance and enable a more natural programming style.
+Empirical evidence also supports this notion that developers prefer first-class support for a parallel abstraction over a library-based solution, if other factors are held constant \cite{ppfirstclass}.
+%
+%Unfortunately, adopting a specialized research language comes with a host of ancillary challenges  \cite{adoptionTR,socioplt}.
+These are significant enough that programming abstractions that are distributed as libraries for a well-established language, such as MapReduce, Global Arrays and MPI, have prevailed in practice, despite the limited expressivity of library-based abstractions.
+This is balanced by several factors: developers can progressively introduce parallelism into existing programs, leverage a familiar and well-developed general-purpose programming language when not directly invoking parallel primitives, and utilize arbitrary combinations of  abstractions within a single program without needing to navigate the friction inherent to language boundaries.
+%
+%Ace aims to bridge the language-oriented and library-oriented approaches by allowing developers to define and distribute a broad range of new {\em first-class} types and their operations directly within user libraries, using a mechanism called {\em active typechecking and translation (\ATT)}.
+Based on the notion of active libraries \cite{activelibraries}, \ATT~is designed to allow developers to  extend the semantics, but not the syntax, of a language.
+The Ace grammar is a nearly-complete subset of Python's  and we show that it is flexible enough to support all of the primitives of the C99 and OpenCL languages in a natural manner, when aided by a form of type inference that we describe.
+We further supplement these primitives with higher-level user-defined primitives, examples of which we discuss in Section \ref{abstractions}.
+Notably, all of these primitives are imported via the normal library loading mechanism; they are not defined within domain-specific languages, nor as extensions for a particular compiler implementation.
+%
+%Although the \ATT~mechanism, a key feature of Ace, is designed primarily for use by researchers and domain experts, Ace is intended to be highly practical and ready for adoption by professional end-user developers today.
+It is based on the Python programming language, which enjoys broad adoption in this community.
+Indeed, the language itself is implemented entirely as a Python library, and many existing tools, such as editor modes, documentation tools and style checkers, can be used without modification.
+%Ace functions are generated programmatically and can be called directly from Python, with numeric arrays as arguments, so developers can integrate Ace immediately within their existing workflows for portions of code that may benefit from static compilation and support for parallelism.
+%We demonstrate the practicality of Ace in Section \ref{clegans} by describing an Ace-based framework that supports specifying and orchestrating families of large-scale neurobiological circuit simulations on clusters of GPUs.
+
+\section{Structure and Usage}
+%Ace is an {\em actively-typed, language-integrated compilation environment}. Let us begin by defining each of these terms. 
+%A {\em compilation environment} is a system that allows developers to programmatically control aspects of compilation directly.
+The language used to do so is called the {\em metalanguage} while the grammar that programs themselves are written in is called the {\em source grammar}.
+The metalanguage of Ace is Python and the source grammar is a significant subset of Python's own grammar.
+Programs and metalanguage code can be interleaved within the same file, so Ace is {\em language-integrated}.
+Ace programs are statically-typed, and types themselves are objects in the metalanguage.
+They can be equipped by developers with methods that control how operators associated with them are typechecked and compiled.
+Because of the relationship between this mechanism and the idea of {\em active libraries} \cite{activelibraries}, discussed further in Section \ref{related}, we call them {\em active types}, and thus Ace is {\em actively-typed}.
+\section{Ace for Researchers}
 \subsection{Use Cases}

### 2013-04-27 ace/ace.tex
Final SC13 submission

@@
 % can use linebreaks \\ within to get better formatting as desired
-\title{Ace: An Actively-Typed Language and Compilation Environment for High-Performance Computing}
+\title{{Ace}: An Actively-Typed Language and Compilation Environment for High-Performance Computing}
 
@@
 The core innovation underlying this and other features of Ace is a novel reification of types as first-class objects at compile-time, representing a refinement to the concept of active libraries that we call \emph{active types}.
-We validate our overall design by considering a case study of a simulation framework enabling the modular specification and efficient execution of ensembles of neural simulations across clusters of GPUs.
+We validate our overall design by considering a case study of a simulation framework enabling the modular specification and efficient execution of ensembles of neural simulations across a cluster of GPUs.
 
@@
 
-We conclude in Sections \ref{related} and \ref{discussion} with related work and a discussion of the limitations of Ace at the time of writing, as well as a discussion of planned future work further validating Ace and building upon the concept of active typing.
+We conclude in Sections \ref{related} and 7 with related work and a discussion of the limitations of Ace at the time of writing, as well as a discussion of planned future work further validating Ace and building upon the concept of active typing.
 
@@
 A variant of the standard ``Hello, World!'' example written in Ace is shown in Listing \ref{hello} and its compilation to statically-typed OpenCL kernel code is demonstrated in Listing \ref{helloout}.
-The OpenCL kernel language is a variant of C99 with some additions and restrictions to facilitate execution on GPUs and other accelerators, in addition to conventional CPUs \cite{opencl}.
+The OpenCL kernel language is a variant of C99 with some additions and restrictions to facilitate execution on GPUs and other accelerators, in addition to conventional CPUs.
 We will discuss it further in subsequent examples.
@@
 This choice was not arbitrary, but rather a key means by which Ace achieves both \textbf{ease-of-use} and \textbf{continuity}.
-Python's whitespace-delimited syntax is widely regarded as both concise and readable, and Python is amongst the most widely-adopted languages in computational science \cite{pythonhpc}.
+Python's whitespace-delimited syntax is widely regarded as both concise and readable, and Python is amongst the most widely-adopted languages in computational science.
 By directly adopting Python's syntax, Ace's syntax is immediately \emph{familiar} and \emph{acceptable} to a significant segment of the intended audience.
@@
 
-We will show in Section \ref{pyopencl} that for backends with Python bindings, such as OpenCL, CUDA and C, generic functions can be executed directly, without this explicit compilation step to concrete functions, if desired.
+We will show in Section \ref{backend} that for backends with Python bindings, such as OpenCL, CUDA and C, generic functions can be executed directly, without this explicit compilation step to concrete functions, if desired.
 
@@
 The \verb|compile| method assigns types to the arguments of a generic function.
-In Listing \ref{mapadd5dbl}, the types we are using are given shoter names, for convenience, on Line 8 (that is, variables in the metalanguage can be used like \verb|typedef| is used in a C-like language).
+In Listing \ref{mapadd5dbl}, the types we are using are given shorter names, for convenience, on Line 8 (that is, variables in the metalanguage can be used like \verb|typedef| is used in a C-like language).
 The types \verb|int| and \verb|double| imported from the \verb|OpenCL| module correspond to the OpenCL types of the same name.
@@
 This is a simple case of Ace's more general {\em whole-function type inference}.
-In this case, \verb|gid| will be inferred to have type \verb|size_t| because that is the return type of \verb|get_global_id| (as defined in the OpenCL specification \cite{opencl}, which the \verb|ace.OpenCL| module follows).
+In this case, \verb|gid| will be inferred to have type \verb|size_t| because that is the return type of \verb|get_global_id| (as defined in the OpenCL specification, which the \verb|ace.OpenCL| module follows).
 The result can be observed on Lines 11 and 24 in Listing \ref{mapout}.
@@
 In addition to type annotations, OpenCL normally asks for additional annotations in a number of other situations.
-Users can annotate functions that meet certain requirements to be callable from the host with the \verb|__kernel| attribute.
-The \verb|OpenCL| backend is able to check these requirements and add this annotation automatically.
-Several types (notably, \verb|double|) and specialized functions require that an OpenCL extension be enabled with a \verb|#pragma| if used.
-The OpenCL backend automatically detects many of these cases as well and adds the appropriate \verb|#pragma| declaration.
-An example of this feature can be seen on Line 5.1, where the use of the \verb|double| type triggers the insertion of an appropriate \verb|#pragma| automatically.
-
-\section{Active Typechecking and\\Translation (AT\&T)}\label{att}
-Thus far, we have been largely discussing the OpenCL module in our examples.
-However, Ace gives no preferential treatment to this module; it is implemented entirely using the user-facing mechanisms described in this section.
-A C99 module has also been developed (which, due to its similarity to OpenCL, shares many of its implementation details), but we do not discuss it further in this paper.
-Extensions complementing these are described in Section C below.
-
-Most programming languages are {\em monolithic} -- the set of available primitives is determined by the language designers, and users must combine these primitives to produce any desired run-time behavior.
-Although many very general primitives have been developed (e.g.
-object systems and algebraic datatypes), these can be insufficient in specialized domains where developers and researchers need fine control over how certain operations are type checked and translated.
-High-performance computing is an example of such a field, since both correctness and performance have been difficult to achieve in general, and are topics of active research.
-As discussed in the Introduction, the proliferation of parallel programming languages, rather than library-based abstractions, indicates that researchers often find the abstractions available in existing languages insufficient for their needs.
-
-To address these use cases, Ace has been designed to be fundamentally {\em extensible}, rather than monolithic.
-Users can introduce new primitive types and operations and fully control how they are typechecked and translated.
-The backend target of translation can also be specified modularly by the user.
-Because Ace libraries can contain compile-time logic, written in the metalanguage as described in the previous section, these primitive definitions can be distributed as modules, rather than as extensions to particular compilers or domain-specific languages.
-
-\subsection{Active Typechecking and Translation (\ATT)}
+Users can annotate functions that meet certain requirements  with the \verb|__kernel| attribute, indicating that they are callable from the host.
+The \verb|OpenCL| backend can check these and add this annotation automatically.
+Several types (notably, \verb|double|) and specialized primitive functions require that an OpenCL extension be enabled via a \verb|#pragma|.
+The OpenCL backend automatically detects many of these cases as well, adding the appropriate \verb|#pragma| declaration.
+An example of this can be seen in Listing \ref{mapout}, where the use of the \verb|double| type triggers the insertion of the \verb|cl_khr_f64| extension.
+
+\section{Extensibility}\label{att}
+Thus far, we have been discussing the \verb|OpenCL| module in our examples.
+This module faithfully implements all the types and operations of the OpenCL kernel language, a portable standard for writing low-level code on multi-core processors and accelerators.
+The functions we wrote have used these primitives with the \verb|OpenCL.OpenCL| backend, so the translation has been direct and no run-time overhead has been introduced.
+Thus, as described so far, Ace has affirmatively resolved the question it was originally conceived to address, discussed in Section 2.1.
+
+\subsection{Monolithic vs. Extensible Languages}
+OpenCL is not necessarily the best tool for every job in high-performance computing.
+Indeed, HPC is an area where designing a set of primitives that satisfy all users has been particularly challenging, and it appears unlikely that a broad consensus will emerge given the variety of architectures, applications, scales and user communities that it serves, and the number of seemingly promising abstractions that emerge continuously targeting various subsets of this problem space.
+
+It is therefore a concern that most programming languages are {\em monolithic} -- a collection of primitives are given first-class treatment by the language implementation, and users can only creatively combine them to implement algorithms and abstractions of their design.
+Although highly-expressive general-purpose mechanisms have been developed (such as object systems or algebraic datatypes), these may not suffice when researchers or domain experts wish to evolve aspects of the type system, exert control over the representation of data, introduce specialized run-time mechanisms, or if defining an abstraction in terms of existing mechanisms is unnatural or verbose (in summary, to push the boundaries of \textbf{verifiability}, \textbf{performance} and \textbf{ease-of-use}).
+In these situations, it would be desirable to have the ability to modularly extend existing systems (\textbf{continuity}) with new compile-time logic (\textbf{extensibility}) and be assured that such extensions will never interfere with one another when used in the same program (\textbf{interoperability}).
+
+%Because Ace libraries can contain compile-time logic, written in the metalanguage as described in the previous section, these primitive definitions can be distributed as modules, rather than as extensions to particular compilers or domain-specific languages.
+\subsection{Active Typechecking and Translation}
+To achieve these criteria, Ace has been designed around a minimal, {extensible} core.
+Users introduce the compile-time logic associated with primitive types and operations from \emph{within libraries}, as opposed to by some language-external mechanism such as a domain-specific language framework or extensible compiler (see Section \ref{related}).
+
+The two phases of compilation that can be controlled by users are together referred to as \emph{active typechecking and translation (AT\&T)}.
+They are invoked the first time the \verb|compile| method of a generic function is called, or when type propagation occurs, with a particular type assignment (a cached concrete function is returned subsequently).
+
+\subsubsection{Active Typechecking}
 \begin{codelisting}
@@
 \end{codelisting}
-We now explain how new primitive types and operators can be defined in Ace.
-When the compiler encounters an expression, such as \verb|input[gid]|, it must first verify its validity by assigning it a type, then translate the expression to produce an expression in a target language.
-Rather than containing fixed logic for this, however, the Ace compiler defers this responsibility to the {\it type} of a subexpression, such as \verb|input|, whenever possible, according to a fixed {\em dispatch protocol} for each syntactic form.
+
+%We now explain how new primitive types and operators can be defined in Ace. 
+When the compiler encounters an expression, it must first verify its validity by either assigning it a type or raising a meaningful type error.
+Rather than defining fixed logic for this, Ace defers control to the {type} of the \emph{primary operand} according to a {\em dispatch protocol}.
+Because types are metalanguage instances of user-defined types derived from \verb|ace.Type| (cf.
+Section \ref{types}), dispatch corresponds to calling a method of this base class.
+
+Let us again consider the data-parallel map example of Section 2.
+When \verb|map| is compiled on Line 9 of Listing \ref{mapadd5dbl}, the first two argument types are \verb|gptr(double)|.
+This type is an instance of the user-defined \verb|OpenCL.PtrType| which inherits from \verb|ace.Type|.
+When the compiler encounters the expression \verb|input[gid]| on Line 6, the dispatch protocol is to defer control to the type of \verb|input| by invoking the method named \verb|resolve_Subscript|.
+
+The relevant portion of \verb|OpenCL.PtrType| as well as \verb|gptr| is shown in Listing \ref{pointers}.
+The \verb|resolve_Subscript| method on Line 8 receives a context and the syntax tree of the node being considered.
+The context contains information about variables in scope and other contextual information, and also exposes a method, \verb|resolve_type|, that this method uses to recursively resolve the types of other subexpressions, here the slice which will be \verb|gid|, as needed.
+The context will have that \verb|gid| is the machine-dependent integer type \verb|size_t| as discussed in Section 2.
+On Line 10, it confirms that this type is an instance of an integer type and thus, it assigns the whole expression, \verb|input[gid]|, the target type of the pointer, \verb|double|.
+Had the function attempted to index \verb|input| using a non-integer expression, the method would take the other branch of the conditional and raise a type error, with a custom error message, on Line 13.
+We note that error messages are an important component of \textbf{ease-of-use} \cite{marceau2011measuring}.
+Indeed, a widely-reported frustration with C++ is that it produces overly verbose and cryptic error messages, particularly when templates are used in clever ways.
+
+\subsubsection{Dispatch Protocol}
 Below are examples of the rules that comprise the Ace dispatch protocol.
@@
 \begin{itemize}
-\item Responsibility over a {\bf unary operation} like \verb|-x| is handed to the type assigned to the operand, \verb|x|.
+\item Responsibility over {\bf unary operations} like \verb|-x| is given to the type assigned to the operand, \verb|x|.
 \item Responsibility over {\bf binary operations} is first handed to the type assigned to the left operand.
-If it indicates that it does not understand the operation, the type assigned to the right operand is handed responsibility, with a different method call\footnote{Note that this operates similarly to the Python run-time operator overloading protocol; see Related Work.}.
-\item Responsibility over {\bf attribute access} (\texttt{obj.attr}) and {\bf subscript access}, (\texttt{obj[idx]}) is handed to the type assigned to \texttt{obj}.
-\item {\color{red} Talk about multiple assignment here I think}
+If it indicates that it does not understand the operation, the type assigned to the right operand is handed responsibility, via a different method call.
+{Note that this operates similarly to the Python run-time operator overloading protocol; see Section \ref{related}.}
+\item Responsibility over {\bf attribute access} (\texttt{e.attr}), {\bf subscript access} (\texttt{e[idx]}) and \textbf{calls} (\verb|e(e1, ..., en)|) is handed to the type assigned to \texttt{e}.
+\item In support of the type inference mechanism described in Section \ref{inference}, responsibility to resolve a type given multiple assignments can be taken by any of the types of the assignments, with priority given to later types.
 \end{itemize}
 
-\subsubsection{Active Typechecking}
-During the typechecking phase, the type of the primary operand, as determined by this dispatch protocol, is responsible for assigning a type to the expression as a whole.
-Let us consider the \verb|map| function from Listing \ref{map} once again.
-When it is compiled on Line 9 of Listing \ref{mapadd5dbl}, its first two argument types are given as \verb|global_ptr(double)|.
-As described in Section \ref{types}, this type, abbreviated \verb|A|, is an instance of \verb|ace.OpenCL.GlobalPtrType| which inherits from \verb|ace.OpenCL.PtrType| and ultimately from \verb|ace.Type|.
-So when the compiler encounters the expression \verb|input[gid]| on Line 6, it follows the dispatch protocol just described and assigns responsibility over typechecking it to \verb|A|.
-This is done by calling the \verb|resolve_|$X$ method of the responsible type, where $X$ is the syntactic form of the expression.
-In this case, the expression is of the \verb|Subscript| form, so the compiler calls \verb|A.resolve_Subscript|.
-
-The relevant portion of \verb|OpenCL.GlobalPtrType| is shown in Listing \ref{pointers}.
-The \verb|verify_Subscript| method on Line 8 receives a context and the syntax tree of the node itself as input.
-The context contains information about other variables in scope, as well as other potentially relevant information, and also contains a method, \verb|resolve_type|, that can be used recursively resolve the types of subexpressions.
-On Line 9, this method is used to resolve the type of the slice subexpression, \verb|gid|, which is the machine-dependent integer type \verb|size_t| as discussed in Section II.E.
-On Line 10, it confirms that this type is an instance of an integer type.
-Thus, it assigns the whole expression, \verb|input[gid]|, the target type of the pointer, \verb|double|.
-Had a user attempted to index \verb|input| using a non-integer value, the method would take the other branch of the conditional and raise a type error with a relevant user-defined error message on Line 13.
-
 \subsubsection{Active Translation}
-Once typechecking a method is complete, the compiler must subsequently translate each Ace source expression into an expression in the target language, OpenCL in the examples thus far.
-It does so by again applying the dispatch protocol and calling a method of the form \verb|translate_|$X$, where $X$ is the syntactic form of the expression.
-This method is responsible for returning a copy of the expression's ast node with an additional attribute, \verb|code|, containing the source code of the translation.
-In this case, it is simply a direct translation to the corresponding OpenCL attribute access (Line 20), using the recursively-determined translations of the operands (Lines 16-17).
+Once typechecking a method is complete, the compiler must subsequently translate each Ace source expression into an expression in the target language.
+This has been OpenCL in the examples thus far, but we describe a generalization of this in the next section.
+
+It does so by again applying the dispatch protocol to call a method of the form \verb|translate_|$X$, where $X$ is the syntactic form of the expression.
+This method is responsible for returning a copy of the expression's ast node with an additional attribute, \verb|code|, containing the source code of the translation, represented here as a string though it may also be represented in a structured manner.
+In our example, it is simply a direct translation to the corresponding OpenCL attribute access (Line 20), using the recursively-determined translations of the operands (Lines 16-17).
 More sophisticated abstractions may insert arbitrarily complex statements and expressions during this phase.
-The context also provides some support for non-local effects, such as new top-level declarations (not shown.)
-
-\subsection{Active Backends}\label{backends}
-Thus far, we have discussed using OpenCL as a backend with Ace.
-The OpenCL extension is the most mature as of this writing.
-However, Ace supports the introduction of new backends in a manner similar to the introduction of new types, by extending the \verb|clq.Backend| base class.
-Backends are provided as the first argument to the \verb|@clq.fn| decorator, as can be seen in Figure \ref{map}.
-Backends are responsible for some aspects of the grammar that do not admit simple dispatch to the type of a subterm, such as number and string literals or basic statements like \verb|while|.
-
-In addition to the OpenCL backend, preliminary C99 and CUDA backends are available (with the caveat that they have not been as fully developed or tested as of this writing.) Backends not based on the C family are also possible, but we leave such developments for future work.
-
-\subsection{Active Translation}\label{translation}
+The context also provides some support for non-local insertions, such as new top-level type declarations, imports and helper code (not shown).
+
+\subsubsection{User-Defined Backends}\label{backends}
+Thus far, we have discussed using OpenCL itself as a backend for our implementation of the OpenCL primitives.
+This backend is called \verb|OpenCL| in the \verb|ace.OpenCL| module.
+Ace supports the definition of new backends in a manner similar to the introduction of new types, by extending the \verb|ace.Backend| base class.
+Backends are specified for a function by using the decorator form \verb|@backend.fn|, as can be seen in the previous examples.
+Backends are responsible for some aspects of the grammar that do not admit simple dispatch to the type of a subterm, such as number and string literals or statement forms like \verb|while| (though \verb|for| is handled by an iterator-like protocol, not shown).
+
+In addition to the OpenCL backend, preliminary C99 and CUDA backends are available (with the caveat that they have not been as fully developed or tested as of this writing.) This allows us to use the OpenCL kernel language, which offers a simplified variant of C99, without relying on the full OpenCL stack, which may not be well-supported by  vendors with proprietary solutions such as CUDA.
+Backends not based on the C family are also possible, but we leave such developments for future work.
+During the translation phase, types can access the backend via the context and emit different code for different supported backends (not shown above due to space considerations).
+
+\subsubsection{Composability and Interoperability}\label{safety}
+Because a type can only exert control over typechecking and translation of operations where an expression of that type is the primary operand, extensions defined using AT\&T can not interfere with one another by construction.
+This does not imply that there are no \textbf{interoperability} issues to consider, however.
+A type may need to know about other types (e.g.
+pointer types need to know about integer types) and if this is done without consideration of future extensions, it may be difficult to integrate data produced by one type system with another.
+These issues cannot easily be addressed by a language design, however.
+
 \subsection{Use Cases}\label{usecases}
-\section{Ace: A Compilation Environment}\label{compenv}
+The development of the full OpenCL language using only the extension mechanisms described above provides evidence of the power of this approach.
+However, to be truly useful, the mechanism must be able to express a wide array of higher-level primitive abstractions.
+We briefly describe a number of other abstractions that are possible using this mechanism.
+Many of these are currently available in existing languages either via libraries or as primitives of some specialized language.
+A study comparing a language-based concurrency solution for Java with an equivalent, though less clean, library-based solution found that language support is preferable but leads to many of the issues we have described \cite{cave2010comparing}.
+
+\subsubsection{Parallel Programming}
+\paragraph{Partitioned Global Address Spaces}
+A number of recent languages designed for clusters use a partitioned global address space model, including UPC, Chapel, X10 and others.
+These languages provide first-class support for accessing data transparently across a massively parallel cluster.
+Their type systems track information about \emph{data locality} so that the compiler can emit more efficient code.
+The extension mechanism of Ace allows inelegant library-based approaches that cannot track such information, such as the Global Arrays library, to be hidden behind a natural array-based notation.
+The locality information can be tracked in a manner analogous to the address space information that is tracked in OpenCL.
+
+\paragraph{Object Models and Message Passing}
+Classes in object-oriented systems can be understood as inducing a type parameterized by a static type specification that emits code to perform dynamic dispatch by some protocol.
+In Ace, the object system would be a subclass of \verb|ace.Type| (e.g.
+\verb|JavaObject|), the specification would be a compile-time Python object (e.g.
+a dictionary mapping names to types, as is done with OpenCL \verb|struct|s, augmented with information about subtyping and methods in more complex systems).
+The code for checking field and method accesses against the specification would be be in \verb|resolve_Attribute| method and the dynamic dispatch mechanism would be emitted in \verb|translate_Attribute|.
+That is, Ace supports arbitrary object models, rather than fixing a single one that may not fit all needs, as in almost all other object-oriented languages.
+
+Many parallel programming abstractions can be understood as implementing object models with complex forms of dynamic dispatch.
+For example, in Charm++, dynamic dispatch involves message passing over a dynamically load-balanced network.
+While Charm++ is a very sophisticated system, and we do not anticipate that implementing it as an Ace extension would be easy, it is entirely possible to do so.
+Erlang and other message-passing based systems can also be considered in this manner -- processes can be thought of as objects, channels as methods and messages as calls.
+
+\paragraph{Functional Parallelism}
+Another promising approach for parallel programming is in automatic parallelization of purely functional programs.
+These languages express computations that do not explicitly manipulate memory.
+They rely instead on the composition of primitives like \verb|map| and \verb|reduce| that transform components of persistent data structures like lists, trees and maps.
+Data dependencies are thus directly encoded in programs, and automatic parallelization techniques have shown promise in using this information to automatically schedule these parallel programs on concurrent hardware and networks.
+
+The Copperhead language, for example, is based on Python as well and allows users to express functional programs over lists using common functional primitives, compiling them to CUDA code.
+We believe that Ace can express precisely the same programming model, implemented as an extension.
+We have prototyped support for a number of common persistent data structures, including algebraic datatypes, in Ace.
+Perhaps surprisingly, Python's imperative syntax does not preclude writing programs in a reasonable functional style (due largely to its support for tuples).
+
+\subsubsection{Other Use Cases in HPC}
+\paragraph{Interoperability Layers}
+The design criteria of \textbf{continuity} and \textbf{interoperability} require consideration of the large body of codes written in a variety of existing languages, across a number of paradigms.
+Although it is often possible to call from one language to another using a foreign function interface (FFI), this is almost never natural or statically safe.
+In Ace, however, extensions could be written that internalize the foreign language's type system (note that dynamically-typed languages can be considered to have a single type, often called $\mathtt{dyn}$) and emit code that hides the FFI from users, achieving \textbf{verifiability} and \textbf{ease-of-use}.
+
+\paragraph{Specialized Optimizations}
+In many cases, specialized code optimizations requires statically tracking invariants throughout a program.
+Often these optimizations can be encoded as a type system for this reason.
+For instance, a course project using Ace  implemented substantial portions of the GPU-specific optimizations described in \cite{yang2010gpgpu} as a library, using types to track affine transformations of the global ID in order to construct a summary of the memory access patterns of the kernel.
+This information can be used both for single-kernel optimization (as in \cite{yang2010gpgpu}) and for future research on cross-kernel fusion and other optimizations.
+
+\paragraph{Domain-Specific Type Systems}
+Although not strictly related to HPC, a number of domain-specific type systems related to computational science can be implemented within Ace.
+For example, prior work has considered tracking units of measure (e.g.
+grams) statically to validate scientific code \cite{conf/cefp/Kennedy09}.
+This cannot easily be implemented using existing abstraction mechanisms because this information should only be maintained statically to avoid excessive run-time overhead associated with tagging.
+The Ace extension mechanism allows this information to be tracked in the type system, but not included during translation.
+
+\paragraph{Instrumentation}
+Several sophisticated feedback-directed optimizations and adaptive run-time protocols require instrumenting code in other ways.
+The extension mechanism supports emitting granular instrumentation that can consider both the syntactic form of an operation as well as its constituent types, easing the implementation of such tools.
+
+This ability could also be used to collect data useful for more rigorous usability and usage studies of languages and abstractions, and we plan on following up on this line of research going forward.
+
+\section{Ace: a Compilation Environment}\label{compenv}\label{backend}
+\begin{codelisting}
+\lstinputlisting{listing8.py}
+\caption{[\texttt{listing\ref{py}.py}] A full OpenCL program using the \texttt{Ace.OpenCL} Python bindings, including data transfer to and from a device and direct invocation of a generic function, \texttt{map}, as a kernel without explicit compilation.}
+\label{py}
+\end{codelisting}
+
+As discussed in the Introduction, a common workflow for professional end-users is to use a high-level scripting language for orchestration, small-scale data analysis and visualization and call into a low-level language for performance-critical sections.
+Python is already widely used by professional end-users in HPC as a high-level scripting language and also features mature support for calling into code written in low-level languages.
+Developers can call into native libraries using its foreign function interface (FFI),  or by using a wrapper library like \verb|pycuda| for code compiled with CUDA or \verb|weave| for C and C++.
+
+Although C, C++ and CUDA's compilers are separate executables on the system, the OpenCL language was designed for this workflow, in that it exposes the compiler and memory management directly as an API, called the \emph{host API}.
+The \verb|pyopencl| module exposes this API and supports basic interoperability with \verb|numpy|, the low-level linear algebra package for Python.
+With both \verb|pyopencl| and \verb|pycuda|, users generate OpenCL or CUDA source code as strings, compile it programmatically, then execute it using the run-time APIs that each library provides \cite{pyopencl}.
+This mode of use can be considered one where Python serves as an \emph{interactive compilation environment}.
+
+Ace supports a refinement to this workflow, as an alternative to the \verb|acec| compiler described above.
+In other words, \verb|acec| can generate source code for kernels but it does not specify how and from what language thet will be called.
+
+At the time of writing, this is only supported with the OpenCL backend, but other backends can implement this interface as well by implementing a simple interface specification.
+The Ace host API wraps and is a superset of the \verb|pyopencl|, adding a simpler type-aware API.
+Both generic functions and concrete functions can then be called like regular Python functions, with additional keyword arguments specifying the global and local size (i.e.
+the number of threads and how they are grouped).
+Because device buffers carry type information, this type information can be propagated from Python to Ace functions directly, as if the call had been within a kernel,  eliminating the \verb|compile| step that we have been using thus far and thus all mention of types.
+
+An example of this for the generic \verb|map| function defined in Listing \ref{map} is shown in Listing \ref{py}, with the call itself on Lines 12-13. 
+The first two arguments to \verb|map| are OpenCL buffers, generated using a simplified wrapper to the \verb|pyopencl| APIs on Lines 9-10.
+This wrapper associates type information with each buffer, based on the type of the \verb|numpy| array, and this is used to implicitly compile \verb|map| as appropriate the first time it is called for any given combination of input types.
+Notice also that \verb|add5| is passed in directly.
+
+By way of comparison, the same program written using the OpenCL C API directly is an order of magnitude larger and significantly more difficult to read and understand.
+It does not support higher-order functions nor is there any way to write \verb|map| in a type-generic way.
+A full implementation of the logic of \verb|map| written using the \verb|pyopencl| bindings and metaprogramming techniques as described in \cite{pyopencl} is still twice as large and more difficult to comprehend than the code we have shown thus far.
+Not shown are several additional conveniences, such as delegated kernel sizing and \verb|In| and \verb|Out| constructs that can reduce the size and improve the clarity of this code further; due to a lack of space, the reader is referred to the language documentation for additional details on these features.
+
+\section{Case Study}\label{casestudy}
+An important criteria that practitioners use to evaluate a language or abstraction is whether significant case studies have been conducted with it, sometimes called \textbf{social proof} \cite{basili2008understanding}.
+In this section, we briefly (due to space constraints) show an application of the mode of use discussed above to support a modular, high-performance scientific simulation library.
+This library was used to simulate  thousands of parallel realizations of a stochastic spiking neurobiological circuit.
+The realizations were distributed over a cluster of GPUs using an existing Python library supporting remote procedure calls.
+This relied critically on Ace's representation of types to determine the memory usage of each realization automatically, and thus distribute realizations and allocate memory memory appropriately.
+It also used metaprogramming features of Ace to generate kernels and specialize them to each node the simulation was running on.
+
 \subsection{Metaprogramming in Ace}
@@
 \lstinputlisting[commentstyle=\color{mauve}]{listing7.py}
-\caption{[\texttt{listing7.py}] Metaprogramming with Ace, showing how to construct generic functions from both strings and abstract syntax trees, and how to manipulate syntax trees at compile-time.}
+\caption{[\texttt{listing9.py}] Metaprogramming with Ace, showing how to construct generic functions from both strings and abstract syntax trees, and how to manipulate syntax trees at compile-time.}
 \label{metaprogramming}
@@
 OpenCL in particular relies on code generation as a fundamental mechanism, which is cited as justification for its lack of support for higher-order programming.
-Ace supports programmatic compilation and higher-order constructs, as described above, and a flexible language extension mechanism, which we describe below, so several use cases for metaprogramming have been eliminated.
-However, cases where this form of metaprogramming include programmatic function specialization (Listing \ref{metaprogramming}) and modular simulation orchestration, described in Section \ref{clegans}.
-
-On Lines 4-7 of Listing \ref{metaprogramming}, an Ace function is constructed from a string containing its source using the \verb|from_source| variant of the \verb|OpenCL.fn| method.
+Ace supports programmatic compilation, higher-order functions and a flexible language extension mechanism, so many use cases for metaprogramming have been eliminated.
+
+However, there are still valid uses for this feature and Ace fully supports it.
+On Lines 4-7 of Listing \ref{metaprogramming}, an Ace function is constructed from a string containing its source using the \verb|from_str| variant of the \verb|OpenCL.fn| method.
 It can also be constructed directly from an abstract syntax tree (AST), as implemented by the Python standard \verb|ast| package, using the \verb|from_ast| variant of \verb|fn|, demonstrated on Line 10.
@@
 This transformation as well as some others are distributed in \verb|ace.astx| for convenience.
-\subsection{Direct Invocation from Python}\label{direct}
-
-\begin{codelisting}
-\lstinputlisting{listing8.py}
-\caption{[\texttt{listing\ref{py}.py}] A full OpenCL program using the \texttt{Ace.OpenCL} Python bindings, including data transfer to and from a device and direct invocation of a generic function, \texttt{map}, as a kernel without explicit compilation.}
-\label{py}
-\end{codelisting}
-
-As discussed in the Introduction, a common workflow for professional end-users involves the use of a high-level scripting language for overall workflow orchestration and small-scale data analysis and visualization, paired with a low-level language for performance-critical sections.
-Python is already widely used by professional end-users as a high-level scripting language and also features mature support for calling into code written in low-level languages.
-Developers can call into native libraries using its foreign function interface (FFI),  or by using a wrapper library like \verb|pycuda| for code compiled with CUDA, a proprietary language similar to OpenCL  specifically targeting nVidia GPU hardware.
-Although CUDA's compilers are separate executables on the system, the OpenCL language was designed for this workflow, in that it exposes the compiler directly as an API.
-The \verb|pyopencl| module exposes this API as well as the OpenCL memory management API to Python.
-With both \verb|pyopencl| and \verb|pycuda|, developers generate source code as strings, compile it programmatically, then execute it using the run-time APIs that each library provides \cite{pyopencl}.
-
-Ace supports this workflow as an alternative to using the \verb|acec| compiler as described above to generate  source code directly from the shell.
-For the OpenCL backend, these bindings are exposed as a wrapper on top of \verb|pyopencl| called \verb|Ace.OpenCL.bindings|.
-Both generic functions and concrete functions written for a backend that supports the direct execution interface (thus far, \verb|OpenCL| and \verb|C99|) can be called like regular Python functions.
-An example of this for the generic \verb|map| function defined in Listing \ref{map} is shown in Listing \ref{py}, with the call itself on Lines 9-10.
-The first two arguments to \verb|map| are OpenCL buffers, generated using a simplified wrapper to the \verb|pyopencl| APIs on Lines 7-8.
-This wrapper associates type information with each buffer, similarly to \verb|numpy|, and this is used to implicitly compile \verb|map| as appropriate the first time it is called for any given combination of input types.
-Explicit calls to the \verb|compile| method, as we have been showing thusfar, are unnecessary if using this method of invocation.
-The final two keyword arguments on Line 10 are parameters that OpenCL requires for execution that determine the number of threads (called the {\em global size}) and thread grouping (the {\em local size}).
-
-By way of comparison, the same program written using the OpenCL C API directly is an order of magnitude larger and correspondingly more complex.
-A full implementation of the logic of \verb|map| written using the \verb|pyopencl| bindings and metaprogramming techniques as described in \cite{pyopencl} is twice as large and significantly more difficult to comprehend than the code we have shown thus far.
-Not shown are several additional conveniences, such as delegated kernel sizing and \verb|In| and \verb|Out| constructs that can reduce the size and improve the clarity of this code further; due to a lack of space, the reader is referred to the language documentation for additional details on these features.
-
-\section{Case Study}\label{casestudy}
-\section{Related Work}\label{related}
-\section{Discussion}\label{discussion}
-%We argue that many of the barriers to adoption faced by new abstractions stem from the fact that new abstractions are too-often developed and distributed together with a new programming language.
-Some prominent cases where such a {\em language-oriented approach} \cite{languageoriented} has been followed include:
-%\begin{itemize}
-%\item UPC, CAF, X10 and Chapel, built around first-class support for a partitioned global address space (PGAS) programming model; 
-%\item Go, Erlang, Charm++, Occam and Axum, built around first-class support for message passing; and
-%\item CUDA, OpenCL, Cilk and NESL, built around first-class support for data parallelism.
-%\end{itemize}
-%
-%The benefits of taking this approach are clear: first-class language, compiler and run-time support for a programming abstraction can improve performance and enable a more natural programming style.
-Empirical evidence also supports this notion that developers prefer first-class support for a parallel abstraction over a library-based solution, if other factors are held constant \cite{ppfirstclass}.
-%
-%Unfortunately, adopting a specialized research language comes with a host of ancillary challenges  \cite{adoptionTR,socioplt}.
-These are significant enough that programming abstractions that are distributed as libraries for a well-established language, such as MapReduce, Global Arrays and MPI, have prevailed in practice, despite the limited expressivity of library-based abstractions.
-This is balanced by several factors: developers can progressively introduce parallelism into existing programs, leverage a familiar and well-developed general-purpose programming language when not directly invoking parallel primitives, and utilize arbitrary combinations of  abstractions within a single program without needing to navigate the friction inherent to language boundaries.
-%
-%Ace aims to bridge the language-oriented and library-oriented approaches by allowing developers to define and distribute a broad range of new {\em first-class} types and their operations directly within user libraries, using a mechanism called {\em active typechecking and translation (\ATT)}.
-Based on the notion of active libraries \cite{activelibraries}, \ATT~is designed to allow developers to  extend the semantics, but not the syntax, of a language.
-The Ace grammar is a nearly-complete subset of Python's  and we show that it is flexible enough to support all of the primitives of the C99 and OpenCL languages in a natural manner, when aided by a form of type inference that we describe.
-We further supplement these primitives with higher-level user-defined primitives, examples of which we discuss in Section \ref{abstractions}.
-Notably, all of these primitives are imported via the normal library loading mechanism; they are not defined within domain-specific languages, nor as extensions for a particular compiler implementation.
-%
-%Although the \ATT~mechanism, a key feature of Ace, is designed primarily for use by researchers and domain experts, Ace is intended to be highly practical and ready for adoption by professional end-user developers today.
-It is based on the Python programming language, which enjoys broad adoption in this community.
-Indeed, the language itself is implemented entirely as a Python library, and many existing tools, such as editor modes, documentation tools and style checkers, can be used without modification.
-%Ace functions are generated programmatically and can be called directly from Python, with numeric arrays as arguments, so developers can integrate Ace immediately within their existing workflows for portions of code that may benefit from static compilation and support for parallelism.
-%We demonstrate the practicality of Ace in Section \ref{clegans} by describing an Ace-based framework that supports specifying and orchestrating families of large-scale neurobiological circuit simulations on clusters of GPUs.
-
-\section{Structure and Usage}
-%Ace is an {\em actively-typed, language-integrated compilation environment}. Let us begin by defining each of these terms. 
-%A {\em compilation environment} is a system that allows developers to programmatically control aspects of compilation directly.
-The language used to do so is called the {\em metalanguage} while the grammar that programs themselves are written in is called the {\em source grammar}.
-The metalanguage of Ace is Python and the source grammar is a significant subset of Python's own grammar.
-Programs and metalanguage code can be interleaved within the same file, so Ace is {\em language-integrated}.
-Ace programs are statically-typed, and types themselves are objects in the metalanguage.
-They can be equipped by developers with methods that control how operators associated with them are typechecked and compiled.
-Because of the relationship between this mechanism and the idea of {\em active libraries} \cite{activelibraries}, discussed further in Section \ref{related}, we call them {\em active types}, and thus Ace is {\em actively-typed}.
-\section{Ace for Researchers}
-\subsection{Use Cases}
-The development of the full OpenCL language using only the extension mechanisms described above provides evidence of the power of this approach.
-Nothing about the core language was designed specifically for OpenCL.
-However, to be truly useful, as described in Sections \ref{multiparadigm} and  \ref{extensions}, the language must be able to support a wide array of primitive abstractions.
-We briefly describe a number of other abstractions that may be possible using this mechanism.
-Many of these are currently available either via inconvenient libraries or in standalone languages.
-With the Ace extension mechanism, we hope to achieve robust, natural implementations of many of these mechanisms within the same language.
-
-\paragraph{Partitioned Global Address Spaces}
-A number of recent languages in high-performance computing have been centered around a partitioned global address space model, including UPC, Chapel, X10 and others.
-These languages provide first-class support for accessing data transparently across a massively parallel cluster, which is verbose and poorly supported by standard C.
-The extension mechanism of Ace allows inelegant library-based approaches such as the Global Arrays library to be hidden behind natural wrappers that can use compile-time information to optimize performance and verify correctness.
-We have developed a prototype of this approach using the C backend and hope to expand upon it in future work.
-
-\paragraph{Other Parallel Abstractions}
-A number of other parallel abstractions, some of which are listed in \ref{multiparadigm}, also suffer from inelegant C-based implementations that spurred the creation of standalone languages.
-A study comparing a language-based concurrency solution for Java with an equivalent, though less clean, library-based solution found that language support is preferable but leads to many of the issues we have described \cite{cave2010comparing}.
-The extension mechanism is designed to enable library-based solutions that operate as first-class language-based solutions, barring the need for particularly exotic syntactic extensions.
-
-\paragraph{Domain-Specific Type Systems}
-Ace is a statically-typed language, so a number of domain-specific abstractions that promise to improve verifiability using types, as discussed in Section \ref{verifiability}, can be implemented using the extension mechanism.
-We hope that this will allow advances from the functional programming community to make their way into the professional end-user community more quickly, particularly those focused on scientific domains (e.g.
-\cite{conf/cefp/Kennedy09}).
-
-\paragraph{Specialized Optimizations}
-In many cases, code optimization requires domain-specific knowledge or sophisticated, parametrizable heuristics.
-Existing compilers make implementing and distribution such optimizations difficult.
-With active libraries in Ace, optimizations can be distributed directly with the libraries that they work with.
-For instance, we have implemented substantial portions of the NVidia GPU-specific optimizations described in \cite{yang2010gpgpu} as a library that uses the extension mechanism to track affine transformations of the thread index used to access arrays, in order to construct a summary of the memory access patterns of the kernel, which can be used both for single-kernel optimization (as in \cite{yang2010gpgpu}) and for future research on cross-kernel fusion and other optimizations.
-
-\paragraph{Instrumentation}
-Several sophisticated feedback-directed optimizations and adaptive run-time protocols require instrumenting code in other ways.
-The extension mechanism enables granular instrumentation based on the form of an operation as well as its constituent types, easing the implementation of such tools.
-This ability could also be used to collect data useful for more rigorous usability and usage studies of languages and abstractions, and we plan on following up on this line of research going forward.
-
-\section{Case Study: Neurobiological Circuit Simulation}
-An important criteria that practitioners use to evaluate a language or abstraction, as discussed in Section \ref{social}, is whether significant case studies have been conducted with it.
-In this section, we briefly (due to space limitations) discuss an application of the Ace OpenCL library, Python host bindings and code generation features for developing a modular, high-performance scientific simulation library used to simulate  thousands of parallel realizations of a spiking neurobiological circuit on a GPU.
 
 \subsection{Background}
-A neural circuit can be modeled as a network of coupled differential equations, where each node corresponds to a single neuron.
-Each neuron is modeled using one or more ordinary differential equations.
-These equations capture the dynamics of physically important quantities like the cell's membrane potential or the conductance across various kinds of ion channels and can take many forms \cite{neurobook}.
+A neural circuit can be modeled as a network of coupled differential equations, where each node corresponds to a single neuron or a small population.
+Each node has temporal dynamics given by one or more ordinary differential equations.
 Single simulations can contain from hundreds to tens of millions of neurons each, depending on the specific problem being studied.
 In some cases, such as when studying the effects of noise on network dynamics or to sweep a parameter space, hundreds or thousands of realizations must be generated.
-In these cases, care must be taken to only probe the simulation for relevant data and process portions of it as the simulation progresses, because the amount of data generated is often too large to store in its entirety for later analysis.
-
-The research group we discuss here (of which the first author was a member) was studying a problem that required running up to 1,000 realizations of a network of between 4,000 and 10,000 neurons each.
+In these cases, care must be taken to only probe the simulation for relevant data and process portions of it as the simulation progresses, because the amount of data generated is often too large to store in its entirety.
+
+The problem we discuss here involved studying a problem that required running up to 1,000 realizations of a network of between 4,000 and 10,000 stochastic neurons each.
 An initial solution to this problem used the Brian framework, written in Python, to conduct these simulations on a CPU.
-Brian was selected because it allowed the structure of the  simulation to be specified in modular and straightforward manner.
+Brian was selected because it allowed the structure of the  simulation to be specified in a modular and straightforward manner.
 This solution required between 60 and 70 minutes to conduct the simulations and up to 8 hours to analyze the data each time a parameter of the simulation was modified.
 
-Unsatisfied with the performance of this approach, the group developed an accelerated variant of the simulation using C++ and CUDA.
+Unsatisfied with the performance of this approach, the author developed an accelerated variant of the simulation using C++ and CUDA.
 Although this produced significant speedups, reducing the time for a simulation by a factor of 40 and the runtime of the slowest analyses by a factor of 200, the overall workflow was also significantly disrupted.
@@
 This quickly led to an incomprehensible and difficult to maintain file structure.
-Moreover, much of the simpler data analysis and visualization was conducted using Python, so marshalling the relevant data between processes also became an issue.
+Moreover, much of the subsequent data analysis and visualization was conducted using Python, so marshalling the relevant data between processes also became an issue.
+Doing so during a simulation was not attempted.
 
 \subsection{The {\sf cl.egans} Simulation Library}
+
+\begin{codelisting}
+\lstinputlisting{listing10.py}
+\caption{\texttt{[listing10.py]} An example of a nested simulation tree, showing that specifying a simulation is both simple and modular.
+The first argument to the constructor specifies each node's parent.}
+\label{spec}
+\end{codelisting}
+
+\begin{codelisting}
+\lstinputlisting{listing11.py}
+\caption{\texttt{[listing11.py]} An example of a hook that inserts code and also inserts new, nested hooks for downstream simulation nodes  below that.}
+\label{impl}
+\end{codelisting}
+
 In order to eliminate these issues while retaining the performance profile of the GPU-accelerated code, the project was ported to Ace.
@@
 {\sf cl.egans} leverages Python's object-oriented features to enable modular, hierarchical simulation specifications.
-For example, Figure \ref{spec} shows an example where a neuron model (\verb|ReducedLIF|) is added to the root of the simulation, a synapse model (\verb|ExponentialSynapse|) is then added to it, and its conductance is probed in the same way, by adding a probe model as a child of the synapse model.
-If interleaved analysis needed to be conducted as well, it would be specified in the same way.
+For example, Figure \ref{spec} shows an example where a neuron model (\verb|ReducedLIF|) is added to the root of the simulation, a synapse model (\verb|ExponentialSynapse|) is then added to it, and its conductance is probed by adding a probe model as a child of the synapse model.
+Interleaved analyses are specified as nodes as well (not shown).
 
@@
 Moreover, because types are first-class objects in the metalanguage, they can be examined during the memory allocation step to enable features like fully-automatic parallelization of multiple realizations across one or more devices, a major feature of {\sf cl.egans} that competing frameworks cannot easily offer.
-
-Once the kernel has been generated and memory has been allocated, the simulation can be executed directly from Python using the bindings described in Section \ref{direct}.
-The results of this simulation are immediately available to the Python code following the simulation and can be visualized and further analyzed using standard tools.
-Once the computations are complete, the Python garbage collector is able to handle deallocation of GPU memory automatically (a feature of the underlying \verb|pyopencl| library \cite{klockner2011pycuda}.)
-
-Using this Ace-based framework, the benefits of the Brian-based workflow were recovered without the  corresponding decrease in performance relative to the previous CUDA-based solution, leading ultimately to a satisfying solution for the group conducting this research.
-
-\begin{codelisting}
-\lstinputlisting{listing10.py}
-\caption{\texttt{[listing10.py]} An example of a nested simulation tree, showing that specifying a simulation is both simple and modular.
-The first argument to the constructor specifies each node's parent.}
-\label{spec}
-\end{codelisting}
-
-\begin{codelisting}
-\lstinputlisting{listing11.py}
-\caption{\texttt{[listing11.py]} An example of a hook that inserts code and also inserts new, nested hooks for downstream simulation nodes  below that.}
-\label{impl}
-\end{codelisting}
-
-\section{Related Work}
-\subsection{Active Libraries in Ace}
-Libraries that have such capabilities has been called {\it active libraries} in prior proposals \cite{activelibraries}.
+Moreover, memory allocation is significantly simplified by the availability of typing information at compile-time.
+The amount of memory needed by a simulation can be calculated statically and thus distributing the simulation over a cluster simply requires determining the available memory on each device in the cluster.
+
+Once the kernel has been generated and memory has been allocated, the simulation can be executed directly  using the bindings above.
+The results of this simulation are available to the Python code during and following the simulation and can be visualized and further analyzed using standard tools.
+Once the computations is complete, the  garbage collector is able to handle deallocation of GPU memory automatically (a feature of the underlying \verb|pyopencl| library \cite{klockner2011pycuda}.)
+
+Using this Ace-based framework, the benefits of the Brian-based workflow were recovered without the  corresponding decrease in performance relative to the previous CUDA-based solution, leading ultimately to a satisfying solution.
+
+\section{Related Work}\label{related}
+\subsection{Active Libraries}
+Libraries that contain compile-time logic have been called {\it active libraries} in prior proposals \cite{activelibraries}.
 A number of  projects, such as Blitz++, have taken advantage of the C++ preprocessor and template-based metaprogramming system to implement domain-specific optimizations.
-In Ace, we replace these brittle mini-languages with a general-purpose language.
-This allows for several interesting uses that we discuss in the following sections.
-
-\subsubsection{Structural Polymorphism}
-In Section \ref{hof}, we discussed several strategies for achieving {\it polymorphism} -- the ability to create functions and data structures that operate over more than a single type.
-In Ace, all functions are implicitly polymorphic and can be called with arguments of {\it any type that supports the operations used by the function}.
-For example, in Figure \ref{map}, \verb|in| can be any type that supports indexing by a variable of type \verb|size_t| to produce a value of a type that can be passed into \verb|fn|, which must then produce a value consistent with indexing into \verb|out|.
-OpenCL pointer types are consistent with these constraints, for example.
-Although powerful, this also demonstrates a caveat of this approach -- that it is more difficult to give a function a concise signature, because arguments are constrained by capability, rather than to a single type \cite{malayeri2009structural}.
+In Ace, we replace these brittle mini-languages with a general-purpose language and significantly expand the notion of active libraries by consideration of types as objects in the metalanguage.
+We thus call these types \emph{active types}.
+
+\subsection{Structural Polymorphism}
+Generic functions represent a novel strategy for achieving {\it function polymorphism} -- the ability to define functions that operate over more than a single type.
+In Ace, generic functions are implicitly polymorphic and can be called with arguments of {\it any type that supports the operations used by the function}.
+This approach is related to structural polymorphism, however \cite{malayeri2009structural}.
+Structural types make explicit the requirements on a function, unlike generic functions.
 
 Structural typing can be compared to the approach taken by dynamically-typed languages that rely on ``duck typing''.
-It is more flexible than the parametric polymorphism found in many functional languages and in languages like Java (which only allow polymorphic functions that are valid for {\it all} possible types), but is of comparable strength to the template system found in C++.
-It can be helpful to think of each function as being preceded by an implicit template header that assigns each argument its own unique type parameter.
-At function call sites, these parameters are implicitly specialized with the types of the provided arguments.
-This choice is again motivated by the criteria of conciseness given in Section \ref{syntax}.
-
-\subsection{Type-Level Computation} %Haskell, Ur and $\Omega$mega
-System XX with simple case analysis provides the basis of type-level computation in Haskell (where type-level functions are called type families \cite{Chakravarty:2005:ATC}).
-Ur uses type-level records and names to support typesafe metaprogramming, with applications to web programming \cite{conf/pldi/Chlipala10}.
-$\Omega$mega adds algebraic data types at the type-level, using these to increase the expressive power of algebraic data types at the expression level \cite{conf/cefp/SheardL07}.
-Dependently-typed languages blur the traditional phase separation between types and expressions, so type-level computation is often implicitly used (though not always in its most general form, e.g.
-Deputy \cite{conf/icfp/ChenX05}, ATS \cite{conf/esop/ConditHAGN07}.)
+It is more flexible than the parametric polymorphism found in many functional languages and in languages like Java (which only allow polymorphic functions that are valid for {\it all} possible types), but is of comparable strength to the template system found in C++ while being much simpler.
 
@@
 MetaML \cite{Sheard:1999:UMS}), and external rewrite systems also exist for many languages.
-These facilities differ from type-level specification in one or more of the following ways:
-
-\begin{enumerate}
-\item In type-level specification, the type of a value is determined separately from its representation; in fact, the same representation may be generated by multiple types. 
-\item We draw a distinction between the metalanguage, used to specify types and compile-time logic, the source grammar, used to describe run-time behavior, and the internal language, used to implement this behavior.
-Term rewriting systems generally do not draw this distinction.
-By doing so, each component language can be structured and constrained as appropriate for its distinct role, as we show.
-%\item With type-level specification, dispatch to a type-level function occurs implicitly on the basis of the structure of an expression.
-In contrast, most term-rewriting systems operate by  explicit invocation of a macro or specialized syntax.
-Some LISP macro systems have explored pattern-based dispatch (e.g.
-A*\cite{fowler2010domain}, EPP\cite{fowler2010domain}) and macro systems for object-oriented languages, like OpenC++ \cite{fowler2010domain} and OpenJava \cite{fowler2010domain}, do offer a somewhat limited form of operation-based dispatch.
-\item Many common macro systems and metaprogramming facilities operate at run-time.
-Compilers for some forms of LISP employ aggressive compile-time specialization techniques to attempt to minimize this overhead.
-Static and staged term-rewriting systems also exist (e.g.
-OpenJava\cite{TatM:OpenJCBMSJ}, Template Haskell\cite{SheardPeytonJones:Haskell-02}, MetaML \cite{Sheard:1999:UMS} and others).
-\end{enumerate}
-
-\subsection{Language Frameworks}
-When the mechanisms available in an existing language prove insufficient, researchers and domain experts must design a new language.
+These differ in their direct exposure to syntax trees and their difficulties with propagating type information, since it is not directly encoded in the syntax.
+The AT\&T mechanism is a type-based mechanism that avoids these issues.
+
+\subsection{Language Frameworks and \\Extensible Compilers}
+When the mechanisms available in an existing language prove insufficient, researchers and domain experts often design a new language.
 A number of tools have been developed to assist with this task, including compiler generators, language workbenches and domain-specific language frameworks (cf \cite{fowler2010domain}).
-
-A major barrier to adoption is the fact that interoperability is intrinsically problematic.
-Even languages which target a common platform, such as the Java Virtual Machine, can only interact using its limited set of primitives.
-Specialized typing rules are not checked at language boundaries, performance often suffers, and the syntax can be unnatural, particularly for languages which differ significantly from the platform's native language (e.g.
-Java).
-
-Instead of focusing on defining standalone languages, type-level specification gives greater responsibility in a granular manner to libraries.
-In this way, a range of constructs can coexist within the same program and, assuming that it can be shown by some method that various constructs are safely composable, be mixed and matched.
-The main limitation is that the protocol requires defining a fixed source grammar, whereas a specialized language has considerable flexibility in that regard.
-Nevertheless, as Ace shows, a simple grammar can be used quite flexibly.
-\subsection{Extensible Compilers}
-An alternative methodology is to implement language features granularly as compiler extensions.
-As discussed in Section 1, existing designs suffer from the same problems related to composability, modularity\-, safety and security as extensible languages, while also adding the issue of language fragmentation.
-
-Type-level specification can in fact be implemented within a compiler, rather than provided as a core language feature.
-This would resolve some of the issues, as described in this paper.
-However, by leveraging type-level computation to integrate the protocol directly into the language, we benefit from common module systems and other shared infrastructure.
-We also avoid the fragmentation issue.
-\subsection{Specification Languages}
-Several {\it specification languages} (or {\it logical frameworks}) based on these theoretical formulations exist, including the OBJ family of languages (e.g.
-CafeOBJ \cite{Diaconescu-Futatsugi01}).
-They provide support for verifying a program against a language specification, and can automatically execute these programs as well in some cases.
-The  language itself specifies which verification and execution strategies are used.
-
-Type-determined compilation takes a more concrete approach to the problem, focusing on combining {\it implementations} of different\- logics, rather than simply their specifications.
-In other words, it focuses on combining {\it type checkers} and {\it implementation strategies} rather than more abstract representations of a language's type system and dynamic semantics.
-In Section 4, we outlined a preliminary approach based on proof assistant available for the type-level language to unify these approaches, and we hope to continue this line of research in future work.
+Extensible compilers can be considered a form of language framework as well due to portability issues that using compiler extensions can introduce.
+It is difficult or impossible for these language-external approaches to achieve interoperability, as discussed above.
+
+\subsection{Extensible Languages}
+Extensible languages like SugarJ \cite{sugarj} afford some of the benefits of Ace, but are not targeted toward Ace.
+They have largely focused on syntactic extensibility, while Ace relies on a fixed syntax, focusing on semantic extensibility instead.
+They also generally allow users to extend languages globally, which leads to conflicts when multiple extensions are used.
+AT\&T does not admit such conflicts by construction.
 
 \section{Conclusion}
-In addition to the novel architecture of Ace as a whole, we note several individually novel features introduced here: %in this paper:
-
-\begin{itemize}
-\item The \ATT~mechanism, which is a generalization of the concept of active libraries \cite{activelibraries}  where types are metalanguage objects.
-\item The method Ace uses to eliminate the need for type annotations in most cases, which combines a form of type inference with type propagation.
-\item The method Ace uses check correctness of generated code by checking representational consistency constraints associated with types, detailed in Section \ref{repcon}. 
-\item The type-aware simulation orchestration techniques used in the \texttt{cl\_egans} library, described in Section \ref{clegans}.
-\end{itemize}
-
-Readers familiar with the Python programming language will recognize the style of syntax used in Figure \ref{map}.
-In fact, Ace uses the Python grammar and parsing facilities directly.
-Several factors motivated this design decision.
-First, Python's syntax is widely credited as being particularly simple and readable, due to its use of significant whitespace and conventional mathematical notation.
-Python is one of the most widely-used languages in scientific computing, so its syntax is already familiar to much of the field.
-And significantly, a large ecosystem of tools already exist that work with Python files, such as code editors, syntax highlighters, style checkers and documentation generators.
-These can be used without modification to work with Ace files.
-Therefore, by re-using an existing, widely-used grammar, we are able to satisfy many of the design criteria described in Section \ref{syntax} and the adoption criteria described in Section \ref{tools} without significant development effort.
 
 Professional end-users demand much from new languages and abstractions.
-In this paper, we began by generating a concrete, detailed set of design and adoption criteria that we hope will be of broad interest and utility to the research community.
-Based on these constraints, we designed a new language, Ace, making several pragmatic design decisions and utilizing advanced techniques, including type inference, structural typing, compile-time metaprogramming and active libraries, to uniquely satisfy many of the criteria we discuss, particularly those related to extensibility.
-We validated the extension mechanism with a mature implementation of  the entirety of the OpenCL type system, as well as preliminary implementations of some other features.
+In this paper, we began by generating a concrete set of design and adoption criteria that we hope will be of interest and utility to the research community.
+Based on these constraints, we designed a new language, Ace, making several pragmatic design decisions and introducing several novel techniques, including type propagation via generic functions, extensible type inference, active typechecking and translation and type-aware Python-Ace-OpenCL bindings to uniquely satisfy many of the criteria we discussed, particularly the three criteria that are typically overlooked in other languages.
+We validated the extension mechanism with a mature implementation of  the entirety of the OpenCL type system, as well as outlined a number of other use cases.
 Finally, we demonstrated that this language was useful in practice, drastically improving performance without negatively impacting the high-level scientific workflow of a large-scale neurobiological circuit simulation project.
-Going forward, we hope that Ace (or simply the key techniques it proposes, by some other vehicle) will be developed further by the community to strengthen the foundations upon which new abstractions are implemented and deployed into professional end-user development communities.
-
-\section{Availability}
-Ace is available under the LGPL license and is developed openly and collaboratively using the popular Github platform at \url{https://github.com/cyrus-/ace}.
+
+Ace has some significant limitations at the moment: debugging is only supported on the generated code, so if code generation introduces significant complexity, this is a significant issue.
+The OpenCL library we have implemented is a reasonably straightforward internalization of OpenCL itself, however, so debugging has not been a problem thusfar.
+We believe that active types can be useful to control debugging, and plan to explore this in the future.
+We will also further explore the use cases and case studies that we have described to validate the design we propose here.
+We hope that Ace will be developed further by the community to strengthen the foundations upon which new abstractions are implemented and deployed into the HPC professional end-user community.
+\newpage
+%\section{Availability}
+%Ace is available under the LGPL license and is developed openly and collaboratively using the popular Github platform at \url{https://github.com/cyrus-/ace}.
 Documentation, examples and other learning materials will be available at \url{http://acelang.org/}.
 {\color{red}(by the time of the conference)}
-
-\section{Acknowledgments}
-CO was funded by the DOE Computational Science Graduate Fellowship under grant number DE-FG02-97ER25308. NF was funded by ...
-
+%
+%\section{Acknowledgments}
+%CO was funded by the DOE Computational Science Graduate Fellowship under grant number DE-FG02-97ER25308. NF was funded by ...
+%
 % An example of a floating figure using the graphicx package.

### 2013-04-27 ace/listing11.py

@@ -1,6 +1,6 @@
-class SpikingModel(Model):
+class ReduceLIF(ModelNode):
   """Base class for spiking neuron models."""
-  def in_model_code(self, g):
+  def in_model_code(self, g): # g is the code generator
     "idx_state = idx_model + count*" << g
     "(realization_n - realization_start)" << g
     self.insert_hook("read_incoming", g)

### 2013-04-27 ace/listing8.py

@@ -1,7 +1,7 @@
 import numpy as np
 import ace.OpenCL.bindings as cl
-from listing1 import map
-from listing2 import add5
+from listing3 import map
+from listing4 import add5
 
 cl.ctx = cl.Context.for_device(0, 0)
 
@@ -13,4 +13,4 @@ map(d_in, d_out, add5,
     global_size=d_in.shape, local_size=(128,))
 
 out = cl.from_device(d_out)
-assert (out == a + 5).all()
+assert (out == input + 5).all()

### 2013-04-27 ace/listing9.py

@@ -1,8 +1,8 @@
-import ace, ace.astx as astx
+import ace
 
 class PtrType(ace.Type):
-  def __init__(self, T, addr_space):
-    self.target_type = T
+  def __init__(self, target_type, addr_space):
+    self.target_type = target_type
     self.addr_space = addr_space
         
   def resolve_Subscript(self, context, node):
@@ -15,12 +15,11 @@ class PtrType(ace.Type):
   def translate_Subscript(self, context, node):
     value = context.translate(node.value)
     slice = context.translate(node.slice)
-    return astx.copy_node(node, 
+    return ace.copy_node(node, 
       value=value, slice=slice,    
-      code=value.code + '[' + slice.code + ']')
-      
+      code=value.code + '[' + slice.code + ']')      
+  
   # ...
-   
-class GlobalPtrType(PtrType):
-  def __init__(self, T):
-    PtrType.__init__(self, T, '__global')
+  
+def gptr(target_type):
+    return PtrType(target_type, "__global")

### 2013-04-28 ace/ace.tex
Final SC13 submission

@@
 
-All examples in Section \ref{usage} are based on an internalization of the OpenCL kernel programming language as a library.
+The examples in Section \ref{usage} are based on an internalization of the OpenCL kernel programming language \cite{opencl11} as a library.
 This library is built atop a novel extensibility mechanism that we call \emph{active typechecking and translation} (AT\&T) and detail in Section \ref{att}.
@@
 The dynamically-typed languages used in HPC, on the other hand, avoid most of this overhead by relying on support from the run-time system.
-Ace was first conceived to explore the question: does conciseness require run-time mechanisms, or can one develop a statically-typed language with the same low-level memory and execution model of C but syntactic overhead comparable to a high-level scripting language?
+Ace was first conceived to explore the question: \emph{does conciseness require run-time mechanisms, or can one develop a statically-typed language with the same low-level memory and execution model of C but syntactic overhead comparable to a high-level scripting language?
+}
 
@@
 This choice was not arbitrary, but rather a key means by which Ace achieves both \textbf{ease-of-use} and \textbf{continuity}.
-Python's whitespace-delimited syntax is widely regarded as both concise and readable, and Python is amongst the most widely-adopted languages in computational science.
+Python's whitespace-delimited syntax is widely regarded as both concise and readable, and Python is amongst the most widely-adopted languages in computational science \cite{oliphant2007python}.
 By directly adopting Python's syntax, Ace's syntax is immediately \emph{familiar} and \emph{acceptable} to a significant segment of the intended audience.
@@
 This significantly simplifies a sophisticated feature of C++ and introduces it to OpenCL and C, which do not support templates.
-Other uses for C++ templates are subsumed by the metaprogramming features discussed in Section \ref{metaprogramming}.
+Other uses for C++ templates are subsumed by the metaprogramming features discussed in Section 5.
 % Is this really any more concise than using C + templates? Etc. relation to types. Conciseness.
@@
 Thus far, we have been discussing the \verb|OpenCL| module in our examples.
-This module faithfully implements all the types and operations of the OpenCL kernel language, a portable standard for writing low-level code on multi-core processors and accelerators.
+This module faithfully implements all the types and operations of the OpenCL kernel language, a portable standard for writing low-level code on multi-core processors and accelerators \cite{opencl11}.
 The functions we wrote have used these primitives with the \verb|OpenCL.OpenCL| backend, so the translation has been direct and no run-time overhead has been introduced.
@@
 Because types are metalanguage instances of user-defined types derived from \verb|ace.Type| (cf.
-Section \ref{types}), dispatch corresponds to calling a method of this base class.
+Section 2.5), dispatch corresponds to calling a method of this base class.
 
@@
 \paragraph{Partitioned Global Address Spaces}
-A number of recent languages designed for clusters use a partitioned global address space model, including UPC, Chapel, X10 and others.
+A number of recent languages designed for clusters use a partitioned global address space model, including UPC \cite{upc}, Chapel \cite{chapel} and others.
 These languages provide first-class support for accessing data transparently across a massively parallel cluster.
 Their type systems track information about \emph{data locality} so that the compiler can emit more efficient code.
-The extension mechanism of Ace allows inelegant library-based approaches that cannot track such information, such as the Global Arrays library, to be hidden behind a natural array-based notation.
-The locality information can be tracked in a manner analogous to the address space information that is tracked in OpenCL.
+The extension mechanism can also track this information in a manner analogous to how address space information is tracked in the OpenCL example above, and target an existing portable run-time such as GASNet \cite{bonachea2002gasnet}.
 
@@
 Many parallel programming abstractions can be understood as implementing object models with complex forms of dynamic dispatch.
-For example, in Charm++, dynamic dispatch involves message passing over a dynamically load-balanced network.
-While Charm++ is a very sophisticated system, and we do not anticipate that implementing it as an Ace extension would be easy, it is entirely possible to do so.
+For example, in Charm++, dynamic dispatch involves message passing over a dynamically load-balanced network \cite{kale1993charm++}.
+While Charm++ is a very sophisticated system, and we do not anticipate that implementing it as an Ace extension would be easy, it is entirely possible to do so using a system like Adaptive MPI that exposes its runtime system \cite{kale2009charm++}.
 Erlang and other message-passing based systems can also be considered in this manner -- processes can be thought of as objects, channels as methods and messages as calls.
@@
 
-The Copperhead language, for example, is based on Python as well and allows users to express functional programs over lists using common functional primitives, compiling them to CUDA code.
+The Copperhead language, for example, is based on Python as well and allows users to express functional programs over lists using common functional primitives, compiling them to CUDA code \cite{catanzaro2011copperhead}.
 We believe that Ace can express precisely the same programming model, implemented as an extension.
@@
 \paragraph{Instrumentation}
-Several sophisticated feedback-directed optimizations and adaptive run-time protocols require instrumenting code in other ways.
-The extension mechanism supports emitting granular instrumentation that can consider both the syntactic form of an operation as well as its constituent types, easing the implementation of such tools.
+Several sophisticated feedback-directed optimizations and adaptive run-time protocols require instrumenting code in various ways.
+The extension mechanism combined with support for patching classes dynamically using Python enables granular instrumentation that can consider both the syntactic form of an operation as well as its constituent types, easing the implementation of such tools.
 
@@
 As discussed in the Introduction, a common workflow for professional end-users is to use a high-level scripting language for orchestration, small-scale data analysis and visualization and call into a low-level language for performance-critical sections.
-Python is already widely used by professional end-users in HPC as a high-level scripting language and also features mature support for calling into code written in low-level languages.
-Developers can call into native libraries using its foreign function interface (FFI),  or by using a wrapper library like \verb|pycuda| for code compiled with CUDA or \verb|weave| for C and C++.
-
-Although C, C++ and CUDA's compilers are separate executables on the system, the OpenCL language was designed for this workflow, in that it exposes the compiler and memory management directly as an API, called the \emph{host API}.
+Python is designed for this style of use \cite{sanner1999python} and is widely used by professional end-users in HPC as a high-level scripting language.
+It features mature support for calling into code written in low-level languages.
+Developers can call into native libraries using its foreign function interface (FFI),  or by using a wrapper library like \verb|pycuda| for code compiled with CUDA \cite{klockner2011pycuda} or \verb|weave| for C and C++.
+
+Although C, C++ and CUDA's compilers are separate executables on the system, the OpenCL language was also designed for this workflow, in that it exposes the compiler and memory management directly as an API, called the \emph{host API}.
 The \verb|pyopencl| module exposes this API and supports basic interoperability with \verb|numpy|, the low-level linear algebra package for Python.
-With both \verb|pyopencl| and \verb|pycuda|, users generate OpenCL or CUDA source code as strings, compile it programmatically, then execute it using the run-time APIs that each library provides \cite{pyopencl}.
+With both \verb|pyopencl| and \verb|pycuda|, users generate OpenCL or CUDA source code as strings, compile it programmatically, then execute it using the run-time APIs that each library provides \cite{klockner2011pycuda}.
 This mode of use can be considered one where Python serves as an \emph{interactive compilation environment}.
@@
 In other words, \verb|acec| can generate source code for kernels but it does not specify how and from what language thet will be called.
-
-At the time of writing, this is only supported with the OpenCL backend, but other backends can implement this interface as well by implementing a simple interface specification.
-The Ace host API wraps and is a superset of the \verb|pyopencl|, adding a simpler type-aware API.
+This mode of use allows Python to be the calling language.
+
+At the time of writing, this is only supported with the OpenCL backend, but other backends can implement this feature as well by satisfying a simple interface specification.
+The OpenCL host API wraps and is a superset of the \verb|pyopencl|, adding a simpler type-aware API.
 Both generic functions and concrete functions can then be called like regular Python functions, with additional keyword arguments specifying the global and local size (i.e.
 the number of threads and how they are grouped).
-Because device buffers carry type information, this type information can be propagated from Python to Ace functions directly, as if the call had been within a kernel,  eliminating the \verb|compile| step that we have been using thus far and thus all mention of types.
+Device buffers can carry type information, unlike in the basic OpenCL host API, and this type information can be propagated from Python to Ace functions directly, as if the call had been within a kernel,  eliminating the \verb|compile| step that we have been using thus far and thus all mention of types.
 
@@
 It does not support higher-order functions nor is there any way to write \verb|map| in a type-generic way.
-A full implementation of the logic of \verb|map| written using the \verb|pyopencl| bindings and metaprogramming techniques as described in \cite{pyopencl} is still twice as large and more difficult to comprehend than the code we have shown thus far.
-Not shown are several additional conveniences, such as delegated kernel sizing and \verb|In| and \verb|Out| constructs that can reduce the size and improve the clarity of this code further; due to a lack of space, the reader is referred to the language documentation for additional details on these features.
+A full implementation of the logic of \verb|map| written using the \verb|pyopencl| bindings and metaprogramming techniques as described in \cite{klockner2011pycuda} is still twice as large and more difficult to comprehend than the code we have shown thus far.
+Not shown are several additional conveniences, such as delegated kernel sizing and \verb|In| and \verb|Out| constructs that can reduce the size and improve the clarity of this code further; due to a lack of space, the reader is referred to the language documentation.
 
@@
 The realizations were distributed over a cluster of GPUs using an existing Python library supporting remote procedure calls.
-This relied critically on Ace's representation of types to determine the memory usage of each realization automatically, and thus distribute realizations and allocate memory memory appropriately.
+This relied critically on Ace's representation of types to determine the memory usage of each realization automatically, and thus distribute realizations and allocate memory on each node appropriately.
 It also used metaprogramming features of Ace to generate kernels and specialize them to each node the simulation was running on.
@@
 Metaprogramming refers to the practice of writing programs that manipulate other programs.
-There are a number of use cases for this technique, including domain-specific optimizations and code generation for programs with a repetitive structure that cannot easily be captured using available abstractions \cite{pyopencl}.
+There are a number of use cases for this technique, including domain-specific optimizations and code generation for programs with a repetitive structure that cannot easily be captured using available abstractions \cite{klockner2011pycuda}.
 OpenCL in particular relies on code generation as a fundamental mechanism, which is cited as justification for its lack of support for higher-order programming.
@@
 Libraries that contain compile-time logic have been called {\it active libraries} in prior proposals \cite{activelibraries}.
-A number of  projects, such as Blitz++, have taken advantage of the C++ preprocessor and template-based metaprogramming system to implement domain-specific optimizations.
+A number of  projects, such as Blitz++, have taken advantage of the C++ preprocessor and template-based metaprogramming system to implement domain-specific optimizations \cite{veldhuizen2000blitz++}.
 In Ace, we replace these brittle mini-languages with a general-purpose language and significantly expand the notion of active libraries by consideration of types as objects in the metalanguage.
@@
 
-Structural typing can be compared to the approach taken by dynamically-typed languages that rely on ``duck typing''.
-It is more flexible than the parametric polymorphism found in many functional languages and in languages like Java (which only allow polymorphic functions that are valid for {\it all} possible types), but is of comparable strength to the template system found in C++ while being much simpler.
+Structural typing can be compared to the more \emph{ad hoc} approach taken by dynamically-typed languages, sometimes called ``duck typing''.
+It is more flexible than the parametric polymorphism found in many functional languages and in languages like Java (which only allow polymorphic functions that are valid for {\it all} possible types), but is comparable to the C++ template system, as discussed previously.
 
@@
 \subsection{Extensible Languages}
-Extensible languages like SugarJ \cite{sugarj} afford some of the benefits of Ace, but are not targeted toward Ace.
-They have largely focused on syntactic extensibility, while Ace relies on a fixed syntax, focusing on semantic extensibility instead.
+Extensible languages like SugarJ \cite{erdweg2011sugarj} afford some of the extensibility benefits of Ace, but are not targeted toward HPC.
+They have largely focused on syntactic extensibility, while Ace relies on a fixed syntax and emphasizes semantic extensibility.
 They also generally allow users to extend languages globally, which leads to conflicts when multiple extensions are used.
-AT\&T does not admit such conflicts by construction.
+AT\&T does not admit such conflicts.
 
@@
 
-Ace has some significant limitations at the moment: debugging is only supported on the generated code, so if code generation introduces significant complexity, this is a significant issue.
+Ace has some limitations at the moment.
+Debugging is only supported on the generated code, so if code generation introduces significant complexity, this can be an issue.
 The OpenCL library we have implemented is a reasonably straightforward internalization of OpenCL itself, however, so debugging has not been a problem thusfar.

### 2013-06-14 globaldsl13/approach.tex
GlobalDSL13 final

@@
+% !TEX root = globaldsl13.tex
 \section{Approach}
 \label{s:approach}
+The examples above demonstrate the core mechanism used in Wyvern: each expression or declaration can contain at most one tilde (\verb|~|).
+The line following the term containing the tilde must begin an indented block.
+This block will be parsed according to a grammar associated with the expected type of the expression where the tilde occurred.
+In Figure \ref{f:dsl-arch}, this type was determined by an explicit type annotation.
+In Figures \ref{f:dsl-query} and \ref{f:dsl-presentation}, it was determined implicitly by the argument types of the function being called.
+Although a single tilde per expression may initially seem limiting, we can see in Figure 3 that the use of a \keyw{where} clause allows for the use of multiple DSLs within a single expression.
 
-Figure~\ref{f:dsl-if} illustrates our approach to type-based extension for the example in Figure 1.
-The class \verb|ComponentArch| defines a grammar extension to the language's expression grammar.
-This grammar must not conflict with the fixed base expression grammar, but because it is only used when an expression of the type \verb|ComponentArch| type is requested \emph{and} a whitespace-delimited block has been introduced, there is no need to guarantee that this grammar does not interfere with other type-associated grammars.
+Figure~\ref{f:dsl-if} shows how users equip types with domain-specific grammars, here for the Architecture example in Figure \ref{f:dsl-arch}.
+The grammar associated with \lstinline{Architecture} is defined at the top level by a specially-named production, \keyw{grammar}, and it also defines two sub-productions for component and policy specifications.
+Each component specification includes a name, a type and an optional list of attributes.
+The name is specified using the \keyw{ID} production, which is a globally-available production that matches valid Wyvern (i.e.
+the GPL's) identifiers.
+Similarly, the \keyw{TYPE} production matches Wyvern types.
+These productions cannot be extended directly, but can be used within DSLs as needed and thus conflicts are detected when the library is compiled, rather than deferred to link-time.
 
-The declared grammar uses a standard BNF notation except for the declaration \verb|EXP of URL| on Line 7.
-This indicates that at this position in the grammar, a term of type \verb|URL| is needed.
-Thus, the domain-specific grammar (if any) associated with the \verb|URL| class may be used here.
-Because URLs are typically short, it may be undesirable to require a new whitespace-delimited block, so the grammar defines blocks delimited by brackets as equivalent to whitespace-delimited blocks as well, as can be seen being used in Figure 1.
+One component attribute in the Architecture DSL is \lstinline{location}.
+On line \ref{f:dsl-if}.6, we see the use of a production defined in another type, \verb|URL|.
+This means that only URL literals are valid at that position, but \emph{not} any Wyvern expression of type \verb|URL|.
+To instead ask for any Wyvern expression of a particular type, we use the notation used in the \lstinline{policy} production on line \ref{f:dsl-if}.8.
+Here, a named policy is defined as an identifier followed by an equals sign and a Wyvern expression of type \lstinline{Policy}.
+This is denoted by the form \keyw{EXP}\verb| : T|, where \verb|T| is a type in scope of the grammar definition.
+Unlike in the URL example above, if there is a DSL associated with this type, it can only be used within a whitespace-delimited block introduced by a tilde.
+This key distinction can be seen in Figure \ref{f:dsl-arch}.
+Again, it can be modularly verified that this grammar does not contain conflicts, assuming that the core Wyvern grammar does not change.
+We expect it to become stable relatively early in its development, with most new features introduced via the embedded DSL mechanism.
 
@@
   \centering\small
-  \begin{lstlisting}[language=Python,morekeywords={grammar,where,ID,TYPE,EXP}]
-class ComponentArch
-  grammar ::= ((component|policy) \n)+ where 
-    component ::= 
-      ("external ")? "component " ID ":" TYPE
-        [componentProperty*]?
-    componentProperty ::= 
-        "location "(EXP of URL)
+  \begin{lstlisting}[language=Python,morekeywords={type,grammar,where,ID,TYPE,EXP}]
+type Architecture
+  grammar ::= (component|policy)+ 
+  component ::= "external"? "component" 
+                     ID ":" TYPE 
+                    ((componentAttr)*)?
+  componentAttr ::= "location " URL.grammar
                        | "connects to" (ID ",")* ID
-    ...
+  policy ::= "policy" ID "=" (EXP : Policy)
   \end{lstlisting}
-  \caption{Type-Associated Grammar Definition}
+  \caption{Type-Associated Grammar}
   \label{f:dsl-if}
@@
 \paragraph{Parsing and Typechecking}
-Wyvern source code is parsed using a standard whitespace-delimited lexer followed by an initial typechecking and parsing phase that occurs in tandem.
+Wyvern source code is parsed in two phases.
+The first phase uses a standard declarative whitespace-delimited parser (e.g.
+\cite{rule2013principled}) where all whitespace-delimited blocks are left as unparsed ``DSL literals''.
+The second phase occurs in tandem with typechecking.
+When a tilde is encountered, the compiler determines the expected type where the tilde appeared (based on function signatures, or explicit type annotations) and parses the subsequent whitespace-delimited block according to the grammar associated with that type.
+The baseline indentation is stripped from this block, so it appears to the parser as if the DSL begins on the leftmost column of the block.
+Any Wyvern expressions that occur internally to a DSL (such as the policy specification in Figure \ref{f:dsl-arch} or curly-brace-delimited expressions in Figure \ref{f:dsl-presentation}) are also parsed and typechecked at this time.
+After parsing a DSL block, it must be verified and translated to Wyvern code.
+The mechanism for doing this is similar, at a high level, to that for defining the grammar itself, but the details of these subsequent steps are beyond the scope of this paper.
+
+\paragraph{Procedural Parsing}
+The grammar definition in Figure 4 is declarative and relies on a parser generator included as part of Wyvern.
+It may be desirable in some cases, such as when a grammar cannot easily be expressed within our declarative framework, or when an existing parser can be called into, to allow a parsing algorithms to be encoded directly.
+An important use case is for interoperability layers between Wyvern and other full-scale programming languages, particularly ones for which parsing is known to be difficult (e.g.
+C, Haskell, Python).
+
+It can be observed that a declarative grammar inside a class definition can be seen as inducing a class method (that is, an operation defined on the class itself, which can be invoked by the compiler or run-time system -- a concept borrowed from Smalltalk \cite{smalltalk}) called \lstinline|parse|, that transforms a string to some AST representation.
+This lower-level interface is exposed directly to programmers who wish to specify parsing in a procedural manner.
+
+%using a standard whitespace-delimited lexer followed by an initial typechecking and parsing phase that occurs in tandem.
 The lexer for Wyvern generates a stream of tokens structured hierarchically, where indentation (and the equivalent bracket notation) increases the depth of the stream.
 Indentation and de-indentation can thus be simply  understood as a special pair of tokens, \texttt{INDENT} and \texttt{DEDENT}, generated by the lexer and delimiting portions of the stream that will be handled in a special way during the next phase.
-
-The standard Wyvern parser reads at a single level of the stream at a time.
+%
+%The standard Wyvern parser reads at a single level of the stream at a time.
 Initially, it parses the stream as a term of sort \verb|Exp|, which represents the core expressions and declaration forms of the language.
@@
 
-When the parsing phase at a particular level is complete, it is typechecked according to a largely conventional protocol.
-In Wyvern, the type of a function is determined by its declaration, and cannot be affected by its use (this constraint may be relaxed in some instances in future iterations of the language).
+%When the parsing phase at a particular level is complete, it is typechecked according to a largely conventional protocol. 
+%In Wyvern, the type of a function is determined by its declaration, and cannot be affected by its use (this constraint may be relaxed in some instances in future iterations of the language).
 Thus, when a function is being applied, the expected types of its arguments are known.
@@
 A delimited block must be present at this location as well, and it will be parsed by the type-associated grammar of $\tau$.
-
-A particular function application can contain only one \emph{primary} indented block.
+%
+%A particular function application can contain only one \emph{primary} indented block.
 If multiple arguments support DSL syntax, then all those other than the first in the argument ordering must be invoked by a partially-indented keyword.
@@
 This can be seen as similar in form to keywords in functions in Smalltalk \cite{smalltalk}.
-
-\paragraph{Keyword-Directed Invocation}
-Keyword-directed invocation of a DSL is simply a special case of this approach.
-In particular, a keyword can be defined as a function with a single argument of a type specific to that keyword.
-The type contains the implementation of the domain-specific syntax associated with that keyword.
-If the construction of this type is controlled such that it can only be via the domain-specific syntax (and not any equivalent syntax in the core language), then it is precisely equivalent to keyword-directed invocation of a DSL.
-
-As keyword-directed DSL invocation is highly explicit, we allow keywords to appear not just at the end of a line, but also within expressions.
-In this case, the keyword determines how much of the expression (and potentially following indented lines) that it will parse.
 
@@
 
-\begin{figure}
-\centering
-\begin{lstlisting}
-db.sendQuery
-    select x where x > 10
-  toServer
-    www.server.org
-\end{lstlisting}
-\caption{Function applications involving multiple white-space delimited arguments use partially-indented keywords.}
-\label{f:dsl-multi}
-\end{figure}
+%\begin{figure}
+%\centering
+%\begin{lstlisting}
+%db.sendQuery
+%    select x where x > 10
+%  toServer
+%    www.server.org
+%\end{lstlisting}
+%\caption{Function applications involving multiple white-space delimited arguments use partially-indented keywords.}
+%\label{f:dsl-multi}
+%\end{figure}
 %\begin{figure}
@@
 %
-\paragraph{Procedural Parsing}
-The grammar definition in Figure 4 is declarative and relies on a simple parser generation framework included in Wyvern.
-It may be desirable for more sophisticated parsing algorithms to be encoded for particular domain-specific languages.
-For example, an interoperability layer with another full-scale programming language may wish to support parsing features not expressible in Wyvern's declarative grammar language.
-It can be observed, however, that a declarative grammar inside a class definition can be thought of as inducing a class method (a concept also borrowed from Smalltalk \cite{smalltalk}) called \verb|parse|, that transforms a stream of tokens to an AST representation.
-This interface can be exposed directly to programmers who wish to extend the parsing capabilities of the language in more sophisticated ways.
-We leave the details of the AST representation and its processing during later stages of compilation as future work.

### 2013-06-14 globaldsl13/conclusion.tex
GlobalDSL13 final

@@
-\section{Conclusion} % and Future Work}
-\label{s:conclusion}
-
-In this paper, we described how extensible parsing in Wyvern makes for
-a solid platform to support whitespace-delimited, type-directed embedded DSLs. In the
-future, we aim to implement a wide variety of DSLs in Wyvern tweaking
-our approach and implementation thereof to provide a comprehensive example of
-supporting multiple interacting DSLs in a safe and easy-to-use manner.
-
+% !TEX root = globaldsl13.tex
+%\section{Conclusion} % and Future Work}
+%\label{s:conclusion}
+%
+%In this paper, we described how extensible parsing in Wyvern makes for
+%a solid platform to support whitespace-delimited, type-directed embedded DSLs. In the
+%future, we aim to implement a wide variety of DSLs in Wyvern tweaking
+%our approach and implementation thereof to provide a comprehensive example of
+%supporting multiple interacting DSLs in a safe and easy-to-use manner.
+%
 % \todo{tie features to goals}
@@
 % \todo{implementation and validation plans}
+
+\section*{Acknowledgements}
+We thank the anonymous reviewers for helpful comments, and acknowledge the support of the Department of Defense and the Air Force Research Laboratory.
+CO is supported by the NSF Graduate Research Fellowship.

### 2013-06-14 globaldsl13/globaldsl13.tex
GlobalDSL13 final

@@
-\documentclass[english]{sigplanconf}
-\usepackage{palatino}
-\usepackage[T1]{fontenc}
-\usepackage[latin9]{inputenc}
+%\documentclass[english]{sigplanconf}
+\documentclass[english]{sig-alternate}
+\usepackage{times}
+%\usepackage[T1]{fontenc}
+%\usepackage[latin9]{inputenc}
 \usepackage{listings}
@@
 \usepackage{graphicx}
+\usepackage[usenames,dvipsnames,svgnames,table]{xcolor}
+\renewcommand{\ttdefault}{pcr}
 %\usepackage{microtype}
@@
 
-\lstset{tabsize=3, basicstyle=\ttfamily\small, commentstyle=\itshape\rmfamily, numbers=left, numberstyle=\tiny, language=java,moredelim=[il][\sffamily]{?},mathescape=true,showspaces=false,showstringspaces=false,columns=fullflexible,xleftmargin=15pt,escapeinside={(@}{@)}, morekeywords=[1]{application,policy,external,component,connects,to,meth,let,in,val,where,return,group,by,within,connect,with,attr,html,head,title,style,body,div,keyword,location,schema,must,salt,select,text}}
+\lstset{tabsize=2, basicstyle=\ttfamily\small, commentstyle=\itshape\rmfamily, numbers=left, numberstyle=\color{gray}\ttfamily\tiny, language=java,moredelim=[il][\sffamily]{?},mathescape=true,showspaces=false,showstringspaces=false,columns=fullflexible,xleftmargin=15pt,escapeinside={(@}{@)}, morekeywords=[1]{application,policy,external,component,connects,to,meth,let,in,val,where,return,group,by,within,connect,with,attr,html,head,title,style,body,div,keyword,location,schema,must,salt,select,text,type,then}}
 \lstloadlanguages{Java,VBScript,XML,HTML,Python}
@@
 
+   \conferenceinfo{GlobalDSL}{'13, Montpellier, France}
+   \CopyrightYear{2013}
+   \crdata{978-1-4503-2043-6}
+
 \begin{document}
 
-\title{Whitespace-Delimited, Type-Directed Parsing for Embedded DSLs}
+\title{Type-Directed, Whitespace-Delimited Parsing for Embedded DSLs}
+\numberofauthors{1} %  in this sample file, there are a *total*
+% of EIGHT authors. SIX appear on the 'first-page' (for formatting
+% reasons) and the remaining two appear in the \additionalauthors section.
+%
+\author{
+% You can go ahead and credit any number of authors here,
+% e.g. one 'row of three' or two rows (consisting of one row of three
+% and a second row of one, two or three).
+% The command \alignauthor (no curly braces needed) should
+% precede each author name, affiliation/snail-mail address and
+% e-mail address. Additionally, tag each line of
+% affiliation/address with \affaddr, and tag the
+% e-mail address with \email.
+% 1st. author
+\alignauthor
+Cyrus Omar, Benjamin Chung, Darya Kurilova, Alex Potanin$^{1}$ and Jonathan Aldrich\\
+       \affaddr{Carnegie Mellon University}\\
+%       \affaddr{1932 Wallamaloo Lane}\\
+%       \affaddr{Wallamaloo, New Zealand}\\
+       \email{\{comar, bwchung, darya, aldrich\}@cs.cmu.edu, and alex@ecs.vuw.ac.nz$^{1}$}
+% 2nd. author
+%\alignauthor
+%G.K.M. Tobin\\
+%       \affaddr{Institute for Clarity in Documentation}\\
+%       \affaddr{P.O. Box 1212}\\
+%       \affaddr{Dublin, Ohio 43017-6221}\\
+%       \email{webmaster@marysville-ohio.com}
+% 3rd. author
+%\alignauthor Lars Th{\o}rv{\"a}ld\\
+%       \affaddr{The Th{\o}rv{\"a}ld Group}\\
+%       \affaddr{1 Th{\o}rv{\"a}ld Circle}\\
+%       \affaddr{Hekla, Iceland}\\
+%       \email{larst@affiliation.org}
+}
 
-\authorinfo{Cyrus Omar, Benjamin Chung$^{1}$, Darya Kurilova, Alex Potanin$^{2}$,
-  and Jonathan Aldrich}
-{Carnegie Mellon University}
-{\{comar, darya, aldrich\}@cs.cmu.edu, bwchung@andrew.cmu.edu$^{1}$, and alex@ecs.vuw.ac.nz$^{2}$}
+%\authorinfo{Cyrus Omar, Benjamin Chung, Darya Kurilova, Alex Potanin$^{1}$ 
+%  and Jonathan Aldrich}
+%{Carnegie Mellon University}
+%{\{comar, bwchung, darya, aldrich\}@cs.cmu.edu, and alex@ecs.vuw.ac.nz$^{1}$}
 % Maybe just keep my cs account active after I leave for a few years?
@@
 
-Domain-specific languages enhance ease of use, expressiveness and
+Domain-specific languages improve ease-of-use, expressiveness and
 verifiability, 
@@
 DSLs within a single application remains difficult.  
-We present an approach to embedded DSLs where:
-1) a whitespace-delimited lexer is shared across DSLs, and 
-2) the parsing and type checking phases occur in tandem so that types can define domain-specific parsing strategies governing whitespace-delimited blocks.
-We argue that this approach occupies a sweet spot with both 
-high expressiveness and ease of use. We outline a design
-and ongoing implementation of this strategy
-and examine the relationship between the type-directed and keyword-directed strategies for parsing of embedded DSLs.
-
+We introduce an approach for embedded DSLs where 1) whitespace delimits DSL-governed blocks, and 2) the parsing and type checking phases occur in tandem so that the expected type of the block determines which domain-specific parser governs that block.
+We argue that this approach occupies a sweet spot, providing   
+high expressiveness and ease-of-use while maintaining safe composability.
+We introduce the design, provide examples and describe an ongoing implementation of this strategy in the Wyvern programming language.
+We also discuss how a more conventional keyword-directed strategy for parsing of DSLs can arise as a special case of this type-directed strategy.
 \end{abstract}
-
 \begin{sloppypar}
@@
 
+\bibliographystyle{abbrv}
+\bibliography{biblio}
 \end{sloppypar}
 
-\bibliographystyle{plain}
-\bibliography{biblio}
 \end{document}

### 2013-06-14 globaldsl13/intro.tex
GlobalDSL13 final

@@
+% !TEX root = globaldsl13.tex
 \section{Introduction}
@@
 
-Domain-specific languages (DSLs) \cite{shellchapter, coldfusion,
-  sinatra}
-% have been widely-studied because they 
-enable the expression of  
-domain-specific ideas naturally and permit domain-specific
-verification and compilation strategies.  For DSLs to reach their full
-potential, however, it must be possible to easily use multiple
-DSLs at once within a host general-purpose language (GPL), such that DSL code
-and GPL glue code can inter-operate to form a complete application. More specifically, we propose the following essential design criteria:
+Domain-specific languages (DSLs) \cite{fowler2010domain} % have been widely-studied because they 
+allow developers to work with   
+specialized abstractions in a natural manner, and allow for specialized 
+verification and compilation strategies that can improve verifiability and performance. However, for DSLs to reach their full
+potential, it must be simple to define a new DSL, invoke it when needed, and to use multiple
+DSLs within a host general-purpose language (GPL), such that pieces of DSL code
+ can interoperate to form a complete application. These intuitions are captured by the following core design criteria that govern our work:
 
@@
 within a single program unit.  %Here 
-Within the text file-based paradigm familiar to most developers, this 
+Within the file-based paradigm used by most contemporary languages, this 
 means including multiple DSLs within a single file.
  Moreover, it should be possible to embed code written in one DSL
-  within another DSL without them requiring knowledge of each other or interfering with one another. So, more specifically, DSLs should be \emph{safely composable}.
+  within another DSL when appropriate, without requiring them to have specific knowledge of each other.
+This should be possible without interference between DSLs used in any combination: DSLs should be \emph{safely composable}.
 
-\item \emph{Interoperability}: DSLs should be able to \emph{interoperate}. A necessary condition for interoperability is that it should be
-  possible to pass and operate on values 
+\item \emph{Interoperability}: It should be
+  possible to pass around and operate on values 
 %such as functions or data structures
-  defined in foreign DSLs. Additional requirements, such as the ability to do so naturally (rather than requiring a large amount of ``glue code'') and safely may also be relevant in many settings. 
+  that were defined in foreign DSLs in a reasonably natural manner (that is, without requiring large amounts of ``glue code'').
+Additional requirements, such as the ability to do so with the safety guarantees provided in the foreign DSL, may also be relevant in many settings.
+
 %Moreover,
-  Minimally, however, if one DSL defines a function or data structure satisfying an interface specified in
-  a common interface description language (such as the type system of
-  a host GPL), code written in another DSL should be able to use those
-  values according to that interface, without requiring that the client DSL
-  have knowledge of the details of the provider DSL or \emph{vice versa}.
-
+  %Minimally, DSLs should be able to define a function or data structure that satisfies an interface specified in
+  %a common interface description language (such as the type system of
+  %a host GPL), code written in another DSL should be able to use those
+  %values according to that interface, without requiring that the client DSL
+  %have knowledge of the details of the provider DSL or \emph{vice versa}.
 \end{itemize}
@@
 
-In addition to composability and interoperability, we believe that to be most useful, a system supporting DSLs should have the following characteristics:
+In addition to these fundamental criteria, we believe that to be most useful, a system supporting DSLs should satisfy the following related design criteria:
 
 \begin{itemize}
-
-\item \emph{Flexibility}: Have flexible enough DSL syntax to support a wide variety of text-based notations;
-
-\item \emph{Identifiability}: Make it easy for programmers to identify which code is written in which DSL;
-
-\item \emph{Consistency}: Encourage DSLs written in similar styles in order to enhance readability and learnability of each DSL;
-
-\item \emph{Simplicity}: Keep the cost of defining a DSL low.
-
+\item \emph{Flexibility}: Support a variety of notations and new language mechanisms, with minimal bias;
+\item \emph{Modularity}: Support defining DSLs as combinations of reusable components distributed directly within  libraries;
+\item \emph{Identifiability}: Make it easy for programmers to identify which code is written in which DSL and what it means;
+%\item \emph{Consistency}: Encourage DSLs written in similar styles whenever possible in order to enhance readability and learnability of each DSL;
+\item \emph{Simplicity}: Keep the complexity and cost of both defining and invoking a DSL as low as possible.
 %\item Share conventions between DSLs and a host language, making each DSL easier for programmers to learn, helping programmers to identify which code is written in which DSL, and avoiding unintended conflicts between DSLs.
 \todo{Avoid conflicts (visual and real), enhance learnability}
-
 %\item Reuse low-level mechanisms and design decisions from a host language, thereby reducing the cost of defining DSLs.  \todo{Easy to define DSLs}
-
 \end{itemize}
 
-We are developing a comprehensive approach that has the potential to meet these design requirements well, based on defining DSLs within a language specifically designed for extensibility from within.
-The approach inherently supports multiple DSLs in the same file, and
-supports interoperability because the DSLs are defined by translation
-into a common internal language.  One key aspect of our proposed solution, and one of the focal points of this paper, is
-that we restrict the syntax of DSLs in a few key ways that greatly simplify
-the mechanism without significantly decreasing expressiveness:
+We are developing a comprehensive language design, \emph{Wyvern}, that we believe can satisfy these design criteria well, and that specifically considers language-internal extensibility from the start.
+In Wyvern, DSL developers define the run-time semantics of DSL constructs via translation
+into a common host language, as in many other DSL frameworks. The novelty of the proposed extensibility mechanism lies in the ways in which we delimit and determine the \emph{scope} of DSL code:
 \begin{itemize}
+%\item The host language and its DSLs share a tokenization and lexing 
+%  strategy, standardizing conventions for identifiers, operators,
+%  constants, and comments.  This avoids the cost of defining lexing
+%  within each DSL, avoids many kinds of low-level clashes between
+%  languages, and makes the composed language more readable. Note that this does not limit the ability of a DSL to define new keywords and other constructs.
+% 
+\item Wyvern is a \emph{whitespace-delimited} language.
+Source code that is governed by a DSL, rather than the GPL, occurs in whitespace-delimited blocks and must be indented further than the GPL line introducing it.
+A decrease in indentation relative to the baseline of the DSL block signals its end.
+This scheme delimits the scope of each DSL in a clear manner, both to
+  the programmer and the top-level parser, supporting the principle of identifiability.
+It also allows Wyvern to avoid restrictions on a DSL's use of delimiters internally.
+Because the GPL grammar is not extended in a global manner, it also guarantees that syntactic
+  conflicts cannot arise at link-time.
 
-\item The host language and its DSLs share a tokenization and lexing 
-  strategy, standardizing conventions for identifiers, operators,
-  constants, and comments.  This avoids the cost of defining lexing
-  within each DSL, avoids many kinds of low-level clashes between
-  languages, and makes the composed language more readable. Note that this does not limit the ability of a DSL to define new keywords and other constructs.
+%There is also a flexible mechanism for explicit delimiters, for small DSLs, that we will discuss briefly later in the paper.
 
-\item Constructs in the host language and its DSLs are
-  \emph{whitespace-delimited} at the top level, according to a particular strategy that we will describe.  Various forms of parentheses 
-  can also serve as delimiters.  Thus, indentation levels or
-  parenthesized expressions clearly delimit blocks that are governed by a particular 
-  language.  This makes the boundaries of each DSL clear to
-  the programmer and the compiler, enhancing usability and guaranteeing that 
-  subtle conflicts cannot arise.
+ % \emph{whitespace-delimited} at the top level, according to a particular strategy that we will describe.  Various forms of parentheses 
+  %can also serve as delimiters.  Thus, indentation levels or
+  %parenthesized expressions clearly delimit blocks that are governed by a particular 
+  %language.  This makes the boundaries of each DSL clear to
+  %the programmer and the compiler, enhancing usability and guaranteeing that 
+  %ubtle conflicts cannot arise.
 
+%\end{itemize}
+
+\item Within this basic syntactic framework, we then propose a novel
+type-directed dispatch mechanism: the \emph{expected type} of an expression, rather than an explicit keyword, 
+%, rather than a keyword, 
+determines which DSL grammar should parse the delimited block that generates that expression.
+That is, \emph{grammars are associated with types}.
+We will show that the  more common keyword-directed strategy arises as a special case of this strategy.
 \end{itemize}
 
-Within this syntactic framework, we then propose a novel syntax extension
-mechanism in which the \emph{expected type of an expression}, rather than an explicit keyword, 
-%, rather than a keyword, 
-signals the switch to a new DSL that parses the tokenized and lexed code within a subsequent delimited block according to a type-associated grammar.
-The  more common keyword-directed strategy arises as a special case of this type-directed strategy.
-We will begin with some motivating examples (\S\ref{s:motivation}), describe our approach (\S\ref{s:approach}) and implementation (\S\ref{s:implementation}), and conclude with related work (\S\ref{s:related}).
+This mechanism allows us to satisfy many of the criteria above, including safe composability, while still being quite expressive, as we will show with examples in the next section.
+We will continue by describing our approach in more detail (\S\ref{s:approach}), discuss ongoing research directions (\S\ref{s:discussion}), and conclude with related work (\S\ref{s:related}).
 

### 2013-06-14 globaldsl13/motivation.tex
GlobalDSL13 final

@@
+% !TEX root = globaldsl13.tex
 \section{Motivating Examples}
@@
   \begin{lstlisting}
-val hotProductDashboard = new Application
+val dashboardArchitecture : Architecture = ~
     external component twitter : Feed
-        location [www.twitter.com]
+        location www.twitter.com
     external component client : Browser
@@
         connects to productDB, twitter
-        location [intranet.nameless.com]
+        location intranet.nameless.com
     component productDB : Database
-        location [db.nameless.com]
-    policy NamelessSecurity
-        must salt [servlet.login.password]
+        location db.nameless.com
+    policy mainPolicy = ~
+        must salt servlet.login.password
         connect * -> servlet with HTTPS
@@
 
-We start with a few examples meant to illustrate the expressiveness of our approach and the breadth of DSLs we plan for it to support.
-The examples are presented in the proposed syntax for Wyvern, a new language being developed by our group  initially targeted toward building secure web and mobile applications.
-We will further clarify how these examples are parsed in Section \ref{s:approach}.
+We start with a few examples to illustrate the expressiveness of our approach and the breadth of DSLs we plan for it to support.
+The examples are presented in the proposed syntax for Wyvern, a new language being developed by our group that is targeted toward building secure web and mobile applications.
+We will informally describe each of these examples here, and further explain how such code is parsed in Section \ref{s:approach}.
 
 The first example, shown in Figure~\ref{f:dsl-arch}, describes the overall architecture of a ``hot product dashboard'' application.
-The constructor \verb|Application| takes an argument of type \verb|ComponentArch|.
-Rather than being provided explicitly, a DSL for specifying the component architecture of the application is used to provide this argument because it appears in a whitespace-delimited block.
-The example architecture declares several components, some of which are marked \keyw{external} to indicate that they are used by this application but are not part of it directly.
-Component types, connectivity and properties, such as location, are declared within further-indented blocks.
-
-The \keyw{policy} keyword introduces a security policy, which constrains the communication protocols to be used as well as
-enforces secure handling of passwords.  Again, indentation delimits
-the scope of the security policy.  This
-%example
-illustrates that one DSL may be nested within another.
+The variable \lstinline{dashboardArchitecture} is explicitly ascribed type \lstinline{Architecture}.
+Rather than explicitly providing a value of this type, we instead use a DSL that makes specifying the component architecture of the application more concise and readable.
+This DSL code appears in the subsequent whitespace-delimited block and is introduced by a tilde (\lstinline{~}).
+The example architecture declares several components, some of which are declared \keyw{external} to indicate that they are used by this application but are not part of it directly.
+Component types are declared after a colon and attributes like connectivity location, are declared after the type (formatted in an indented block for readability).
+The \keyw{policy} keyword (line 11) introduces a security policy, which constrains the communication protocols that can be used and 
+enforces the secure handling of passwords.
+A separate type, \lstinline{Policy}, is associated with such policies.
+Although we could instantiate this type explicitly using a Wyvern expression, we use a DSL for defining policies instead, again within a whitespace-delimited block introduced by a tilde.
 
@@
   \begin{lstlisting}
-val newProds = productDB.query 
+val newProds = productDB.query(~)
     select twHandle 
     where introduced - today < 3 months
-this.prodTwt = Feed(newProds)
-return prodTwt.query
+val prodTwt = new Feed(newProds)
+return prodTwt.query(~)
     select *
     group by followed
-    where count > POPULAR  
+    where count > 1000
   \end{lstlisting}
@@
 
-Figure~\ref{f:dsl-query} shows how a DSL for querying can be used from within ordinary Wyvern code.
-The example shows the body of a method for computing a feed that is derived from tweets about the company's new products.
-In this example, the use of a querying DSL is triggered by the use of a function, query, expecting a \verb|DBQuery|.
-This type defines a particular syntax for database queries, delimited by the indented block as the reader would expect.
-This is similar to what can be expressed with built-in query syntax such as Microsoft's LINQ \cite{mslinq}, but in this case, defined in a user-defined type.
-The feed query operator may define different features, as seen on Line 5.
-\begin{figure}
+Figure~\ref{f:dsl-query} shows how a DSL for database queries can be used from within ordinary Wyvern code.
+The example shows code for computing a feed that is derived from tweets about a company's new products.
+In this example, the use of a querying DSL is triggered by the use of methods named \lstinline{query} expecting an argument of type \lstinline{DBQuery} (line 1) or \lstinline{FeedQuery} (line 5) respectively.
+These types define related but distinct syntax for queries, determined by the expected type of expression where the tilde appears (tildes need not appear only at the ends of lines).
+Queries are again delimited by indentation.
+This mechanism is similar to what can be expressed in languages with built-in query syntax like LINQ \cite{mslinq}, but in this case, it is entirely user-defined, rather than built into the language.\begin{figure}
   \centering
   \begin{lstlisting}
+serve(page, loc) where 
+  val page = ~ 
     html:
       head:
-        title: [text "Hot Products"]
-        style:
-            myStylesheet
+        title: Hot Products
+        style: {myStylesheet}
         body:
           div id="search":
-            SearchBox("products")
+            {SearchBox("products")}
           div id="products":
-            FeedBox(servlet.hotProds())
+            {FeedBox(servlet.hotProds())}
+  val loc = ~
+    products.nameless.com
   \end{lstlisting}
-  \caption{Wyvern DSL: Presentation}
+  \caption{Wyvern DSLs: Presentation and URLs}
   \label{f:dsl-presentation}
@@
 
-Finally, Figure~\ref{f:dsl-presentation} shows a DSL for rendering the hot product application in a web browser.
-The DSL is based on HTML but uses our indentation-based syntax and allows Wyvern code to be integrated -- function/method calls and variables of the appropriate type can be used at any point.
-Literal tags are introduced by a tag name followed by any tag attributes followed by a colon.
+Finally, Figure~\ref{f:dsl-presentation} shows a DSL for presenting the hot product application to a web browser, served at a particular URL.
+Here, two DSLs are used within a single function call.
+To allow this without introducing ambiguity, the user can use a \keyw{where} clause, similar to that found in Haskell \cite{jones2003haskell}.
+The presentation DSL is based on HTML and associated with a type, \lstinline{HTMLElement}.
+It uses an indentation-sensitive syntax and allows integration of Wyvern code of the appropriate type using curly braces.
+The second DSL simply canonicalizes URL literals into Wyvern values of type \lstinline{URL}.

### 2013-06-14 globaldsl13/related.tex
GlobalDSL13 final

@@
+% !TEX root = globaldsl13.tex
 \section{Related Work}
 \label{s:related}
-The most common mechanism for creating programming language extensions is macros, as exemplified by hygienic macros in Scheme.
-Macros in Scheme and other Lisp-style languages are written in Lisp itself, and benefit from the regular syntax of the underlying language and the consistent use of parentheses as expression delimiters.
-The syntactic nature of macros allows them to control parsing within a delimited scope, and this delimited control of parsing inspired our approach in Wyvern.
-Wyvern's use of types to trigger custom parsers avoids the Lisp limitation of starting each DSL with a keyword or macro name.
-
-Wyvern chooses whitespace delimitation rather than parenthesis delimitation because, anecdotally, many programmers find syntax in that style to be more readable.
-There are several proposals for making Lisp and its dialects whitespace-delimited, including sweet-expressions \cite{sweet-expr} and SRFI-49 \cite{srfi-49} for Scheme, P4P proposal for Racket \cite{p4p-proposal}, Kernel's f-expressions \cite{fexprthesis} for Common Lisp \cite{f-expr}, and others.
-None of them seem to be successfully implemented and deployed within the Lisp communities, however.
-This may be because Lisp programmers don't want to switch away from parentheses, yet having an alternate syntax is not by itself enough to draw non-Lisp programmers into the community.
+The most well-known mechanism for extending languages is macros, as exemplified by hygienic macros in Scheme.
+Macros in Scheme and other Lisp-style languages are written in the language itself and benefit from its simple syntax -- parentheses universally serve as expression delimiters (although proposals for whitespace as a substitute for parentheses have been made \cite{srfi-49}).
+Our work is inspired by this flexibility, but aims to support richer syntax as well as static types.
+Wyvern's use of types to trigger parsing  avoids the overhead of needing to invoke macros explicitly by name and makes it easier to compose DSLs declaratively.
 
@@
 
-Some language extensibility projects provide extension at levels of abstraction above parsing. For instance, OJ (previously, OpenJava)
+Some language extensibility projects provide metaprogramming facilities at levels of abstraction above parsing. For instance, OJ (previously, OpenJava)
 \cite{Tatsubori00openjava:a} provides a macro system based on a meta-object
-protocol, and Backstage Java~\cite{Palmer:2011:BJM:2048066.2048137} employs compile-time meta-programming.
-Both of these systems provide macro-style rewriting of existing source code, but they provide at most limited extension of language parsing.
+protocol, and Backstage Java~\cite{Palmer:2011:BJM:2048066.2048137}, Template Haskell \cite{sheard2002template} and others employ compile-time meta-programming.
+Each of these systems provide macro-style rewriting of source code, but they provide at most limited extension of language parsing.
 
-Other systems aim at providing more flexible forms of syntax extension compared to our whitespace-delimited approach, at the potential cost of complexity or conflicts.
-Camlp4 \cite{camlp4} is a preprocessor for OCaml that offers the developer the ability to extend the concrete syntax of the language via the use of parsers and extensible grammars.
-SugarJ \cite{Erdweg:2011:SLL:2048147.2048199} takes a library-centric approach which allows to syntactically extend the Java language by adding libraries.
-In Wyvern, the core language (that is, the \verb|Exp| sort) does not get extended directly, so conflicts cannot arise and reasoning about which DSL is applicable is much simpler.
+Other systems aim at providing forms of syntax extension that change the base language, as opposed to our whitespace-delimited approach.
+For example, Camlp4 \cite{camlp4} is a preprocessor for OCaml that offers the developer the ability to extend the concrete syntax of the language via the use of parsers and extensible grammars.
+SugarJ \cite{Erdweg:2011:SLL:2048147.2048199} takes a library-centric approach which supports syntactic extension of the Java language by adding libraries.
+In Wyvern, the core language (particularly the \keyw{EXP} sort) is not extended directly, so conflicts cannot arise at link-time.
 
-Finally, researchers have developed a number of DSL frameworks and workbenches, such as the Meta Programming System (MPS) \cite{mps}, Spoofax \cite{KatsVisser2010}, the Intentional Domain Workbench (IDW) \cite{idw}, and Ens\={o} \cite{enso}, that provide support for extending a variety of programming languages by applying generic rules and concepts.
-The Marco language \cite{lee:2012:marco} similarly provides macro definition at a level of abstraction that is largely independent of the target language.
+Scoping DSLs to expressions of a single type comes at the expense of some flexibility, but we believe that many uses of DSLs are of this form already.
+A previous approach has considered type-based disambiguation of parse forests for supporting quotation and anti-quotation of arbitrary object languages \cite{bravenboer2005generalized}.
+Our work is similar in spirit, but does not rely on generation of parse forests and associates grammars with types, rather than types with grammar productions.
+We believe that this is a more simple and flexible methodology.
+The remaining approaches to syntax extension, such as XJ \cite{DBLP:conf/scam/ClarkSW08} are keyword-directed in some form.
+We believe that a type-directed approach is more seamless and general, sacrificing a small amount of identifiability in some cases.
 
-Compared to these approaches, Wyvern focuses on making one programming language extensible with embedded DSLs, provides a common lexer and extensible parsing confined to whitespace-delimited regions, and investigates a novel type-based approach to triggering parser extensions.
+Researchers have also developed DSL frameworks and language workbenches, including MPS \cite{mps}, Spoofax \cite{KatsVisser2010}, Ens\={o} \cite{enso} and others \cite{krahn2008monticore,van1992pregmatic} that provide support for generating new programming languages and tooling in a modular manner.
+%The Marco language \cite{lee:2012:marco} similarly provides macro definition at a level of abstraction that is largely independent of the target language.
+Compared to these approaches, Wyvern focuses on extensibility \emph{internal} to the language, rather than taking an approach where each DSL is \emph{external} relative to the host language, improving interoperability and composability.
+
+Finally, recent work on Active Code Completion is related to this work in that it associates code completion palettes with types \cite{omar2012active}.
+Such palettes could be used for defining a DSL syntax for types.
+However, that syntax is  immediately translated to Java syntax at edit-time while this work integrates with the core parsing facilities of the language.
 

### 2013-08-07 syzygy/syzygy.tex
Added syzygy draft

\documentclass{article} % For LaTeX2e
\usepackage[usenames,dvipsnames,svgnames,table]{xcolor}
\usepackage{nips12submit_e,times}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{url}
\usepackage{color}
\usepackage{graphicx}
%\documentstyle[nips12submit_09,times,art10]{article} % For LaTeX 2.09

\title{Structured Statistical Source Code Prediction}
\author{
  Cyrus Omar\\
  \texttt{comar@cs.cmu.edu}
  \And
  Salil Joshi\\
  \texttt{salilj@cs.cmu.edu}
  \And Fl\'avio Cruz\\
  \texttt{fmfernan@cs.cmu.edu}
}

\newcommand{\fix}{\marginpar{FIX}}
\newcommand{\new}{\marginpar{NEW}}

\usepackage{xypic}

\nipsfinalcopy % Uncomment for camera-ready version

\begin{document}

\maketitle

\section*{Introduction}
Programming languages are formal systems with rich syntactic and semantic structure. They are
also human systems, in that they are used extensively by people in patterned ways to express their
intent. Many tools are designed to help people write code more efficiently by predicting
the source code that a developer intends. For example, code completion systems for editors like Eclipse for Java display pop-up menus containing the members of to the class of the variable being manipulated.

Typical code completion systems use only this sort of semantic
information about the language and the libraries being used, and do not incorporate data about how developers have written programs
in the past. Recent work by Hindle et al.~\cite{Hindle:2012:NS:2337223.2337322} demonstrated, however, that source code could be successfully predicted statistically
using a simple $n$-gram model that used a tokenized, rather than structural, representation of source code.

Our project aims to unite the structured and statistical approaches to source code prediction.
That is, rather than using a tokenized representation of source code, we would like to do
statistical prediction on a more natural representation of source code: the typed syntax tree.  We can then condition our predictions using semantic information, specifically:
\begin{itemize}
\item the \emph{type}, denoted $\tau$, of the expression being predicted (e.g. \verb|int| or \verb|Color|)
\item the \emph{syntactic context}, denoted $\sigma$, in which the expression occurs (e.g. whether the expression is an argument of a function call, the guard of an \verb|if| statement, etc.)
\item the \emph{program context}, denoted $\Gamma$, in which the expression occurs (i.e. the set of variables paired with their types that are in scope at the location that the expression occurs.)
\end{itemize}

For example, if a user has entered the code \texttt{Planet destination = }, where \verb|Planet| is an enumeration containing \verb|Mercury|, \verb|Venus|, \verb|Earth|, etc. (but not \verb|Pluto|, of course), then we have that the expected type at the cursor is \verb|Planet|, the syntactic context is \emph{assignment}, and given a program context, our prediction space should only assign non-zero probabilities to:
\begin{itemize}
\item literal members of the \verb|Planet| enumeration (e.g. \verb|Planet.EARTH|)
\item variables and fields of type \verb|Planet| available from the program context
\item calls to methods available from the program context that have return type \texttt{Planet}\footnote{We can consider  operators like $+$ and $[]$ as methods of the built-in types in Java.}
\end{itemize}
To assign a probability to an expression, denoted $e$, we first determine how likely it is that the expression is of each of these three \emph{syntactic forms}, where syntactic forms are denoted $\phi$. For each form, we can then assign probabilities to particular expressions of that form according to some form-specific conditional distribution. This model can be understood as a graphical model, as shown in Figure \ref{graphicalmodel}, where the syntactic form is a latent variable. The conditional distributions for both the syntactic form and expression are learned using data gathered from analyses of prior code corpuses (smoothed using some suitable method in cases where enough information is not available).

\begin{figure}
    \begin{displaymath}
    \xymatrix{
      *+[F]{\color{ForestGreen}\txt{Type ($\tau$)}} \ar[dr] \ar[drr] & & \\
      & *+[F]{\color{black} \txt{Syntactic\\Form ($\phi$)}} \ar[r] & *+[F]{\color{cyan}\txt{Expression ($e$)}}  \\
      *+[F]{\color{ForestGreen}\txt{Syntactic\\Context ($\sigma$)}} \ar[ur] \ar[urr] & & *+[F]{\color{ForestGreen}\txt{Program\\Context ($\Gamma$)}} \ar[u]  
    }
    \end{displaymath}
\caption{A graphical model representing our approach. Green variables are always observed (we do not assign marginal distributions to them.) The syntactic form, $\phi$, is a latent variable, and the expression, $e$, is unknown. Note that the syntactic form is a function of the expression. \label{graphicalmodel}}
\end{figure}

In addition to the applications to code completion systems in code editors like Eclipse, more accurate source code prediction techniques could be useful for other programming tools. For example, programmers with severe physical impairments may benefit from predictive programming interfaces that allow them convey source code using devices more limited than a keyboard, incorporating the predictions to reduce the amount of information that needs to be conveyed explicitly by the user.

\section*{Related Work}
As mentioned above, work by Hindle et al. \cite{Hindle:2012:NS:2337223.2337322} showed that using a simple $n$-gram model of source code, represented as sequences of tokens, produced reasonably accurate predictions. This work (published this summer) suggested improving this model with structural and semantic information as future work. We have corresponded with the authors of this paper.

Work examining the ``uniqueness'' of software has found that source code can often be decomposed into commonly-seen motifs \cite{Gabel:2010:FSE}, further suggesting that prediction that uses structural information may be a particularly helpful approach in the domain of source code.

Code completion systems that use statistical information in a limited context, such as specifically for predicting which methods of a class will be called, have been proposed in the literature \cite{Bruch:2009:LEI:1595696.1595728,robbes_how_2008}. Other systems, such as Calcite \cite{mooty_calcite:_2010}, have produced contextually-relevant {\em code snippets} or examples drawn from a code corpus or web search, where more frequently encountered snippets are preferentially shown. These can all be construed as related, more specialized, variants of the approach we are taking in our project.

\section*{Method}
\newcommand{\asgn}{\textbf{assignment}}
\newcommand{\statement}{\textbf{statement}}
\newcommand{\argument}{\textbf{arg}}
\newcommand{\other}{\textbf{other}}

\newcommand{\lit}{\textbf{lit}}
\newcommand{\meth}{\textbf{meth}}
\newcommand{\var}{\textbf{var}}

\newcommand{\nm}[1]{\#\{#1\}}
\newcommand{\nmm}[1]{\#_M\{#1\}}

\subsection*{Probability Model}
      Our goal is to learn a model that allows us to produce $\mathbf{P}(e|\tau, \sigma, \Gamma)$. As we described informally above, we determine this probability by marginalizing over a latent variable that represents the \emph{syntactic form} of the expression, denoted $\phi$:
    $$P(e | \tau, \sigma, \Gamma) = \sum_{\phi \in \Phi} P(e | \phi, \tau, \sigma, \Gamma) P(\phi | \tau, \sigma)$$
As diagrammed in Figure \ref{graphicalmodel}, $\tau$ is the type of the expression, $\sigma \in \Sigma$ is the syntactic context, $\Gamma$ is the program context, and $\phi \in \Phi$ is the syntactic form. We consider syntactic contexts $\Sigma = \{\statement, \asgn, \argument, \other\}$ representing plain statements, assignments to variables, method arguments and a catch-all for other syntactic contexts (e.g. conditional and loop guards, return statements, etc.). We consider syntactic families $\Phi = \{\lit, \var, \meth\}$, representing literals (of built-in and enumeration types), variables and method calls, respectively.

	  In the equations below, we use the following notation to represent quantities learned from training data:
	  \begin{itemize}
	  \item $\nm{e, \tau, \phi, \sigma}$ represents the number of expressions in the training set constrained by the provided expression, type, syntactic form and program context (summing over any omitted categories.)
	  \item $\nmm{\tau_0.m, \sigma}$ represents the number of method calls to the method $\tau_0.m$ in the training set, constrained by the provided program context.
  	  \end{itemize}

  The conditional distribution for the syntactic forms is categorical, with the probability for each $\phi \in \Phi$ learned as:
  $$P(\phi | \tau, \sigma) = 
  \left\{
  	\begin{array}{ll}
	 	\frac{\nm{\phi,\tau, \sigma}}{\nm{\tau, \sigma}} & \nm{\tau, \sigma} \neq 0\\
		\frac{\nm{\phi, \sigma}}{\nm{\sigma}} & \mbox{o/w}
	\end{array}
	\right.
  $$

  That is, we use the empirical probability that that the expression has syntactic form $\phi$ given that the expression has type $\tau$ and is in syntactic context $\sigma$. If we have no data for type $\tau$ in that syntactic context, we marginalize over $\tau$.

  The conditional distribution for an expression $e$ given the syntactic form, if the actual syntactic form of $e$ is $\phi_e$, is broken down according to the syntactic form as follows:

$$P(e | \phi, \tau, \sigma, \Gamma) =
\left\{
	\begin{array}{ll}
		P_{\lit}(e | \tau, \sigma, \Gamma) & \mbox{if } \phi = \phi_e = \lit\\
		P_{\var}(e | \tau, \sigma, \Gamma) & \mbox{if } \phi = \phi_e = \var\\
		P_{\meth}(e | \tau, \sigma, \Gamma) & \mbox{if } \phi = \phi_e = \meth\\
		0  & \mbox{if } \phi \neq \phi_e 
	\end{array}
\right.
$$

We now describe how probabilities for expressions of each of the three syntactic forms are computed in the following sections.

\paragraph{Literals}
For literals, we use the empirical probability of previously seen literals, assigning 0 probability to all others. Although ideally we would assign non-zero probability to unseen literals, this is complex for literals of built-in types, and not doing so simplifies our implementation considerably, so we leave it as future work (discussed below).

$$P_{\lit}(e | \tau, \sigma, \Gamma) = 
\left\{
	\begin{array}{ll}
		\frac{\nm{e, \lit, \tau, \sigma}}{\nm{\lit, \tau, \sigma}} & \nm{\lit, \tau, \sigma} \neq 0\\
		0 & \mbox{o/w}
	\end{array}
\right.$$

\paragraph{Variables} Variables have highly limited lifespans, so there is little empirical data that we can gather about their usage within a novel context. Thus, we simply give all variables available in the program context equal probability. We denote the set of variables of type $\tau$ in $\Gamma$ as $V(\Gamma, \tau)$.

$$P_{\var}(e | \tau, \sigma, \Gamma) = 
\left\{
	\begin{array}{ll}
	\frac{1}{|V(\Gamma, \tau)|} & e : \tau \in \Gamma\\
	0 & \mbox{o/w}
	\end{array}
\right.$$
$$V(\Gamma, \tau) = \{ x | x : \tau \in \Gamma\}$$

\paragraph{Methods} For method invocations, we first assign a probability for the method itself, and then recursively assign probabilities to the argument expressions. The product of these probabilities is thus the probability of the expression as a whole.

$$P_{\meth}(e | \tau, \sigma, \Gamma) = 
\left\{
	\begin{array}{ll}
          P_m(\tau_o.m | \tau, \sigma, \Gamma) \prod_{i=1}^n P(e_i | \tau_i, \argument, \Gamma)  & \mathcal{M}(e, \tau, \Gamma)\mbox{ true} \\
	0 & \mbox{o/w}
	\end{array}
\right.$$

We must check that the expression is in fact a well-typed method invocation, which we can do using a type theoretic predicate:

$$\begin{array}{lll}
\mathcal{M}(e, \tau, \Gamma) & = & \exists o, m, n, e_1, \ldots, e_n, \tau_o, \tau_1, \ldots, \tau_n. (e \equiv o.m(e_1, \ldots, e_n))~\land\\
 & & 
\Gamma \vdash o : \tau_o~\land~\Gamma \vdash o.m : \tau_o \to (\tau_1 \to \cdots \to \tau_n) \to \tau~\land~\\
 & & \Gamma \vdash e_1 : \tau_1 \land \cdots \land \Gamma \vdash e_n : \tau_n
\end{array}$$

To calculate the probability of a method itself, we consider two cases. It may be a method that we have seen in the training set, in which case we use the empirical probability of that method in the provided syntactic context. It may also be a method that hasn't been seen in the training set. In this case, we give a uniform distribution over all such methods callable via a variable in the program context, which we denote $M(\Gamma, \tau)$. We weigh each of these two possibilities by approximating the likelihood that an unseen method is used by the empirical probability that a method was used at least once in the training set, with a penalty factor $\eta$ in the denominator to ensure that some probability remains for unseen methods. 
\begin{align*}
P_m(\tau_o.m | \tau, \sigma, \Gamma) & =  P_{\text{unseen}}(\tau, \sigma)\frac{1}{|M(\Gamma, \tau)|} + 
P_{\text{seen}}(\tau, \sigma)\frac{\nmm{\tau_o.m, \sigma}}{\nm{\meth, \tau, \sigma}}\\
P_{\text{unseen}}(\tau, \sigma) & =  \frac{ |\{\tau_o.m | \nmm{\tau_o.m, \sigma} > 0\}|}{\nm{\meth, \tau, \sigma} + \eta}\\
P_{\text{seen}}(\tau, \sigma) & =  1 - P_{\text{unseen}}(\tau, \sigma)
\end{align*}

\subsection*{Implementation}
We used the Java programming language to perform experiments to analyze the effectiveness of our methodology. The ideas underlying our project are applicable to any statically typed programming language, but Java has some advantages that are are relevant to us. First, there are large Java code corpuses freely available. Second, there is prior work on predicting Java programs that we can use as a basis for evaluating our model \cite{QualitasCorpus:APSEC:2010}. Lastly, we can use the Eclipse IDE which gives us access to a powerful toolchain for working with Java programs.

We implemented our prediction model using the Eclipse Java Development Tools (JDT) library. This library provides facilities for parsing Java code, for traversing the AST to visit the relevant parts of the tree (such as method declarations or variable bindings etc.) and for retrieving type information.

To learn the parameters of our model, we determine the frequencies of each expression form, given its type and also it's syntactic context.
As above, the context can be either a \textit{variable declaration} (expression appears as the initializer of a variable), \textit{statement}, \textit{method argument} (expression is the argument of a method invocation) or \textit{other} (everything else).

We have implemented three code visitors to collect data about the three expression forms:

\begin{description}
   \item[Literal Visitor:] For primitive types such as integers, floats, doubles or strings we count the frequency of each literal in the code. For enumeration types (like \texttt{Planet}), we count the frequency of each element of the enumeration.

   \item[Variable Visitor:] For each type, we count the number of times a variable of this type was referenced in the code.

   \item[Method Visitor:] We group methods by the return type, so that we can count the number of times each method is used. We also collect data about all the declared methods.
\end{description}

For prediction, we make use of another visitor in order to go through all the expressions in the target file. To compute the probability of an expression $e$ at a certain point in the file, we determine the context $\sigma$ of $e$ and it's type $\tau$. We
then count the number of literals, variables and methods that we found during the collection phase with the same $\sigma$ and $\tau$. We can then do a case analysis on the form of $e$, consistent with the formulation given above:

\begin{itemize}
   \item For literals, we compute the relative frequency of all literals over all expression forms, and multiply it by the relative frequency of the literal value in relation to all literals with the same type and context.
   \item For variables we also compute the relative frequency of variables over all expression forms and divide this by the number of variables of the same type in the scope of $e$. To find all variables, we first find the method $m$ where $e$ is located and then traverse the AST of $m$ to locate all variables that appear before $e$ and have type $\tau$.
   \item For method invocations, we compute the relative frequency of methods over all expression forms and multiply it by the probability of a method being seen or not seen depending  on whether it was found during the data collection phase. If the method was not found before, we find all unused methods with the same return type in the current scope. Finally, we recursively multiply the previous result by the probability of each argument of the method $e$, to get the prediction of the whole expression.
\end{itemize}

\section*{Experiments}

%%Although the best method to test the quality of our models would be to perform an user study, we are a bit more limited in time and resources. Therefore, we used already available software code to automatically test our prediction facilities.

Our experiments use some of the freely available code corpuses listed in Qualitas Corpus~\cite{QualitasCorpus:APSEC:2010}.
Table~\ref{tab:projects} presents the projects we picked from the catalogue. We also show the number of files and
lines of code (LOC) of every project.

\setlength{\textfloatsep}{25pt}
\begin{table}[t]
\centering
\begin{tabular}{|l|c|c|l|}
\hline
\textbf{Project} & \textbf{\# Files} & \textbf{\# LOC} & \textbf{Description} \\
\hline
Ant & 1196 & 256039 & a software tool for automating software build processes \\
ANTLR & 394 & 66845 & a framework for implementing compilers \\
Art Of Illusion & 487 & 153645 & an open source 3D modelling and rendering studio \\
Axion & 237 & 42329 & a relational database system supporting SQL and JDBC \\
Batik & 1664 & 366506 & a library for manipulation of SVG files \\
FindBugs & 1149 & 207827 & a program that uses static analysis to look for bugs \\
%JFreeChart & 1005 & 321551 & a library for displaying professional charts \\
\hline
\end{tabular}
\caption{Software projects used in our experiments.}
\label{tab:projects}
\end{table}
For each project, we consider a subset of files for testing, the test set, and all the remaining files as the training set.
In our experiments we want to determine the probability of an observed expression $e$ according to our model (the prediction rate). Ideally, $P(e| \tau, \sigma, \Gamma)$ should be 1 since $e$ was the expression that was actually observed. Thus a high prediction rate implies an accurate model.

To reduce noise in our prediction rate, we use a ten-fold cross validation, where we select 10\% of the files at random as the test set and build our model by using the remaining 90\% of the files. For each file in the test set, we visit all the expressions (including sub-expressions) and compute the prediction rate. We then average the prediction rate of all expressions found to produce the prediction rate for that file. Finally, we average the prediction rate of all files to get a prediction rate for the current iteration of the cross validation procedure.

\subsection*{Baseline: The $n$-gram model}
For comparison with the state-of-the-art, we generated probabilities using an $n$-gram model~\cite{Hindle:2012:NS:2337223.2337322}, with $n = 3$ tokens. The $n$-grams were generated using a version of the CMU Language Modeling Toolkit and the probabilities were computed using a Python scripts as follows:

$$P(t_3, \ldots, t_n | t_1, t_2) = \prod_{i=3}^n P_{\text{ngram}}(t_i | t_{i-2}, t_{i-1})$$

In the above, the tokens $t_3, \ldots, t_n$ are the tokens from the expression, $e$, and $t_1$ and $t_2$ are the tokens immediately preceding $e$ in the program.

\subsection*{Results}
Figure \ref{results} compares the probabilities generated by both methods, showing that our method (SSCP) performed significantly better than the $n$-gram model across all projects.
\begin{figure}
\begin{centering}
\includegraphics[scale=0.4]{overall.png}
\end{centering}
\caption{Results showing the average probability assigned to the actual expression for both the $n$-gram model (red) and our model (blue) for each of the six projects that we considered.\label{results}}
\end{figure}

\begin{figure}
\begin{centering}
\includegraphics[scale=0.4]{ant.png}
\end{centering}
\caption{A histogram showing the distribution of probabilities assigned to expressions in the Ant project. The bucket for probabilities below 0.1 is omitted, as explained in the text. \label{hist}}
\end{figure}

We also analyzed the probabilities generated in a few of our runs to try and find areas where we can improve. Figure \ref{hist} shows a histogram of probabilities generated from one run on the ant project. The histogram is shown without the bucket of probabilities $< 0.01$ since there are so many of them. Analysis of our data revealed the following facts:

\begin{itemize}
  \item Our predictions are very good for variables. The average probability is high and even probabilities above 0.9 are not uncommon. This indicates that variables are used in predictable ways.
  \item Our predictions are very bad for \texttt{String}, \texttt{float}, \texttt{double}, and to a certain extent \texttt{int} literals. This is not surprising given our method, but these literals are so common that it would be useful to do something smarter in these cases. For example, for \texttt{int} literals we could imagine using an exponentially decreasing distribution that makes smaller numbers more probable, with added weights given to numbers we have seen before. For \texttt{Strings} we could use an underlying language model.
  \item Our predictions are also not very good when the expression is a method call. However this is only to be expected. In the case of a method call we are predicting a much larger expression, since we recursively predict the arguments and then multiply the probabilities. Our results improve dramatically if we consider only the \emph{name} of the method being called, which is all that would be required in an IDE code completion setting.
\end{itemize}

We also performed tests across projects i.e.\ we trained on one project and then calculated probabilities for expressions in a different project. The results are given in Table \ref{tab:cross}.

\begin{table}[h]
\centering
\begin{tabular}{|c|c|c|c|c|c|c|c|}
\hline
           & Ant   & ANTLR & AoI   & Axion & Batik & FindBugs \\
\hline
 Ant       & -     & 14.44 & 12.19 & 11.39 & 12.54 & 12.96    \\
ANTLR      & 14.04 & -     & 11.33 & 10.17 & 10.64 & 10.84    \\
AoI        & 18.94 & 17.3  & -     & 17.12 & 16.77 & 15.92    \\
Axion      & 15.93 & 14.74 & 12.08 & -     & 11.85 & 11.9     \\
Batik      & 18.71 & 17.35 & 15.11 & 13.84 & -     & 13.94    \\
FindBugs   & 17.23 & 16.00 & 14.16 & 12.97 & 13.18 & -        \\
\hline
\end{tabular}
\caption{Cross Project probabilities.}
\label{tab:cross}
\end{table}

From this table we find that in most cases, within-project accuracy is higher than cross-project accuracy, which means that our method does learn facts that are local to the project (AoI is an exception to this rule, and we do not yet have a good explanation for its behavior).

Further, it seems possible that we can use cross-project prediction to learn interesting facts about the similarities between projects. For example, Ant is bad at predicting other projects, except for ANTLR. This is perhaps not a coincidence, because the Ant project has been designed to work well with ANTLR and has specific support for it. However, more analysis is required before we can be can determine whether cross-project accuracy is meaningful.

\section*{Conclusion}

We developed a prediction model for source code completion through the use of type and context information. Upon testing our model on open source code corpuses, we found that our model is several times more accurate than the current state of the art \cite{Hindle:2012:NS:2337223.2337322}, which indicates that using language semantics in addition to statistical information is useful.

There are several avenues for future work:
\begin{itemize}
  \item In the previous section we identified some shortcomings of our current model and discussed some possible improvements. We are actively working on these.
  \item We would like to use our model to provide better code completion in Eclipse, and as an aid to limited input devices.
  \item It seems possible that cross-project predictions can provide information about the similarities between projects. We would like to investigate this idea further.
\end{itemize}

\section*{Acknowledgments}

We are grateful to YoungSeok Yoon for providing example code for an Eclipse plug-in that we used to get started,  to Rachel Aurand for providing source code for the CMU Language Toolkit, and to Avinava Dubey for helpful feedback on the midway report.

\bibliography{midway}
\bibliographystyle{plain}

\end{document}

### 2013-08-16 thesis-proposal/problem.tex
Updated thesis statement

@@
-\section{Introduction}
+\section{Introduction}\begin{figure}
+\begin{center}
+\includegraphics[scale=0.5]{approaches.pdf}
+\end{center}
+\vspace{-20px}
+\caption{\small (a) With the language-oriented approach, novel constructs are packaged into separate languages.
+Users can only safely and naturally call into languages consisting of common constructs.
+(b) With the library-oriented approach, there is one language and novel constructs are packaged as normal libraries.
+Thus, interoperability is not a problem but in monolithic systems, this approach is less expressive.\label{approaches}}
+\end{figure}
+Designing and implementing a programming language together with its supporting tools (collectively, a \emph{programming system}) that has a sound theoretical foundation, supports a natural programming style, helps users identify and fix errors as early as possible, and performs well across diverse problem domains remains a grand challenge in computing.
+No small, fixed collection of abstractions and associated tools that can adequately address these criteria in all situations.
+Instead,  researchers and domain experts regularly design and implement novel domain-specific abstractions, provide alternative implementations of existing abstractions, and supplement these implementations with a variety of edit-time, compile-time and run-time behaviors in order to better satisfy these criteria within the constraints of particular application domains.
 
-Jonathan Aldrich (ISR), chair
-Brad A. Myers (HCII)
-Robert Harper (CSD)
-Premkumar Devanbu (UCSD)
+To realize a new abstraction or behavior, researchers and domain experts must take either a \emph{language-internal approach}, where they work entirely within an existing language and distribute their solutions as libraries, or a \emph{language-external approach}, where they either construct a new, distinct system or extend an existing system by some mechanism that is not part of the language itself (such as an extension mechanism supported by a particular compiler or development environment).
 
-Designing and implementing a programming language together with its supporting tools (collectively, a \emph{programming system}) that simultaneously supports a natural programming style, helps users find and fix errors early, and consistently performs well across diverse application domains remains a grand challenge in computing.
-Indeed, these concerns are often at odds with one another, so system designers must make trade-offs.
-No small set of general-purpose constructs can completely satisfy every need.
-Evidence supporting this view comes from the large community of researchers and domain experts designing new domain-specific languages and tools, each tailored to the needs of a particular developer community or application domain.
+When it is possible, taking a \emph{language-internal approach} (aka a \emph{library-oriented approach}), where new constructs are implemented entirely within a single host language as libraries, is a far simpler and more practical approach.
+% -- library-based constructs cannot interfere with one another (assuming a suitable namespacing standard), libraries benefit from common distribution infrastructure and tool support, and externally-accessible library interfaces need not be restricted to a fixed common subset of all available constructs (e.g.
+the object system of the JVM or CLI).
+We diagram this key latter distinction in Figure \ref{approaches}.
+Of course, this is very often \emph{not} possible today because from the perspective of a library developer, the programming system is a largely inflexible, \emph{monolithic} entity.
+That is, the language syntax and its static and dynamic semantics are fixed in advance, the compiler is a ``black box'' implementation of these fixed semantics, and other tools operate according to domain-agnostic protocols that use only the basic structure of the program being developed as input.
+To enhance any of these components of the programming system, the expert must step out of the language, and often out of the programming system entirely.
 
-%In implementing a new feature or abstraction, these experts can take one of several approaches. 
-To realize a new construct, these experts often take a \emph{language-oriented approach}, developing and distributing it together with a new programming language.
+Thus, for the most sophisticated new constructs and behaviors, experts often prefer the  {language-oriented approach}, developing and distributing it together with a new programming language.
 The flexibility of this approach, coupled with support from tools like compiler generators and language workbenches, has led to a proliferation of specialized languages, each built around a particular set of privileged constructs.
 Unfortunately, a key issue limiting the practicality of this approach occurs at the boundaries between languages -- specialized constructs particular to one language cannot always be safely and naturally expressed in another, so building a program out of components written in different specialized languages can often prove difficult.
+
+{\color{red} (Need to reorganize this bit a little bit)}
+
+Although it is widely recognized that typed programming languages are well-positioned to achieve these goals, there is no small collection of types that can adequately satisfy these criteria in all cases.
+Instead, a large community of researchers and domain experts continue to design new domain-specific type systems and implement them in various programming systems, making trade-offs between these concerns as appropriate for the needs of the research or developer community they are targeting.
 
@@
 
-Stepping back, it is clear that a \emph{library-oriented approach}, shown in Figure \ref{approaches}b, where new constructs are implemented entirely within a single host language as libraries, should be preferred when possible.
-% -- library-based constructs cannot interfere with one another (assuming a suitable namespacing standard), libraries benefit from common distribution infrastructure and tool support, and externally-accessible library interfaces need not be restricted to a fixed common subset of all available constructs (e.g.
-the object system of the JVM or CLI).
-We diagram this key latter distinction in Figure \ref{approaches}.
-Of course, this is very often \emph{not} possible today because from the perspective of a library developer, the programming system is a largely inflexible, \emph{monolithic} entity.
-That is, the language's syntax, type checking rules and semantics are fixed in advance, the compiler is a black box implementation of these rules, and the tools operate according to domain-agnostic protocols that use only the declared structure of the program being developed as input.
-To enhance any of these aspects of the programming system, the expert must step out of the language, and often out of the programming system entirely.
-
 To put it another way, monolithic programming languages enforce a fundamental dichotomy between \emph{first-class constructs}, which enjoy support throughout the programming system but must be designed and implemented by the system designer in advance, and \emph{user-defined constructs}, which can be distributed as user libraries, but must creatively combine and repurpose the small set of available first-class constructs to express all desired run-time, compile-time and edit-time behaviors.
@@
 But this then couples their core innovations with a collection of other, unrelated design choices, requires more effort both for the experts developing the innovation and its targetted user community, and leads to the intrinsic interoperability problems described above for users developing applications consisting of components written in different languages.
-
-\begin{figure}
-\begin{center}
-\includegraphics[scale=0.5]{approaches.pdf}
-\end{center}
-\vspace{-20px}
-\caption{\small (a) With the language-oriented approach, novel constructs are packaged into separate languages.
-Users can only safely and naturally call into languages consisting of common constructs.
-(b) With the library-oriented approach, there is one language and novel constructs are packaged as normal libraries.
-Thus, interoperability is not a problem but in monolithic systems, this approach is less expressive.\label{approaches}}
-\end{figure}

### 2013-08-16 thesis-proposal/timeline.tex
Updated thesis statement

@@
 
-Our plan for carrying out the work described above involves building significantly on existing work on Ace, as well as iterating on existing highly-simplified prototypes we have developed of all of the components described above.
+\subsection{Spring 2013}
+I will complete the work on Ace and active type theory and submit it for publication.
 
-In the first year, we will fully develop the theory of active types and completing development of the Ace language.
-We plan on releasing Ace to the broader community, as described in Section 5.1, within the coming year.
-In addition, we will implement initial versions of a core active code prediction system for Java.
+\subsection{Summer 2013}
+I will complete the work on active code prediction and submit it for publication.
 
-In the second year, we will begin conducting capstone evaluations of Ace, further applying the theory to functional languages, and begin working on applications of active code prediction to code completion and rehabilitative programming systems.
-
-In the third year, we will complete capstone evaluations of Ace, conduct capstone evaluations of active code prediction and the applications that we have developed, and begin prototyping an actively-typed structured editor, laying foundations for the long-term evolution of this research program beyond the scope of this grant.
-
-Evaluations will consist of user studies, case studies, classroom studies and field studies based on instrumented versions of the tools that we develop.
-We plan on justifying not just the theoretical utility of active types, but their usefulness and usability in practice as well.
+\subsection{Fall 2013}
+I will finalize all work and write the final dissertation.

### 2013-08-16 thesis-proposal/work.tex
Updated thesis statement

@@
 
-\subsection{Syntax and Visual Representation}
-Thus far, we have been discussing active types within the context of languages with a fixed syntax.
+%\subsection{Syntax and Visual Representation}
+%Thus far, we have been discussing active types within the context of languages with a fixed syntax.
 However, in many specialized domains, a specialized syntax or visual representation for novel constructs may be appropriate.
@@
 We have not settled on any one of these, so our proposed research will include a rigorous examination of the pros and cons of each approach.
-
-\subsubsection{Extensible Parsing} 
-Extensible parsers have been well-studied in the literature (e.g.
+%
+%\subsubsection{Extensible Parsing} 
+%Extensible parsers have been well-studied in the literature (e.g.
 \cite{journals/entcs/BrabrandSV03}), and such systems could be integrated with an actively-typed language without conflict with one caveat: each new syntactic form needs to have a dispatch protocol associated with it, to assign control over its semantics to one of its subexpressions.
 Such a mechanism will be one avenue we will evaluate for suitability.
-
-\subsubsection{Flexible Reinterpretation}
-An approach that we have prototyped, but not extensively utilized, in Ace is to allow active types to directly observe the syntax trees of subexpressions, rather than simply their types as we have described above.
+%
+%\subsubsection{Flexible Reinterpretation}
+%An approach that we have prototyped, but not extensively utilized, in Ace is to allow active types to directly observe the syntax trees of subexpressions, rather than simply their types as we have described above.
 This allows a highly contextualized reinterpretation of a syntax tree, similar to capabilities found in macro systems (although without a direct rewriting component.) We will further explore this approach, particularly as we begin implementing abstractions that may not naturally fit within Python's existing syntactic forms in an intuitive way.
-
-\subsubsection{Delimited Parsing}
-A related alternative approach would be to partially parse a file, leaving delimited regions either unparsed or only tokenized and dispatching to an active type to do the actual parsing.
+%
+%\subsubsection{Delimited Parsing}
+%A related alternative approach would be to partially parse a file, leaving delimited regions either unparsed or only tokenized and dispatching to an active type to do the actual parsing.
 This would give more flexibility to the active type, but also a greater responsibility to fully parse and give semantics to the provided source code.
 Nevertheless, this may be appropriate in certain cases where a highly-specialized local syntax is desirable for convenience (and flexible reinterpretation or active code completion at edit-time does not suffice.)
-
-\subsubsection{Actively-Typed Structured Editors}
-Projects such as Citrus \cite{Ko05} and Barista \cite{Ko06} are analogs of language frameworks for structured editors -- editors where source code elements are represented directly as a tree in a visual or semi-visual form, rather than written as a string as in conventional languages.
+%
+%\subsubsection{Actively-Typed Structured Editors}
+%Projects such as Citrus \cite{Ko05} and Barista \cite{Ko06} are analogs of language frameworks for structured editors -- editors where source code elements are represented directly as a tree in a visual or semi-visual form, rather than written as a string as in conventional languages.
 These tools demonstrate that modular construction of such editors may be possible.

### 2013-08-26 thesis-proposal/conclusion.tex
Updated abstract, most of introduction, beginning of approach

% !TEX root = omar-thesis-proposal.tex
\section{Conclusion}

Monolithic programming systems enforce a dichotomy between \emph{built-in constructs}, which enjoy support throughout the programming system but must be designed and implemented by the system designer in advance, and \emph{user-defined constructs}, which can be distributed as user libraries, but must creatively combine and repurpose the small set of available built-in constructs to express all desired run-time, compile-time and edit-time behaviors. This dichotomy has forced researchers and domain experts who want to significantly advance the state-of-the-art in programming systems, particularly in compiled, statically-typed systems where this dichotomy is most salient, to design and develop new languages. But this then couples their core innovations with a collection of unrelated design decisions, requires more effort both for the experts developing the innovation and its targeted user community, and leads to intrinsic problems for users developing applications consisting of components written in different languages.
\newpage
\listoftodos
\newpage

### 2013-08-26 thesis-proposal/problem.tex
Updated abstract, most of introduction, beginning of approach

--- 2013-01-09 thesis-proposal/problem.tex
@@
-\section{Problem}
+% !TEX root = omar-thesis-proposal.tex
+\section{Introduction}\begin{figure}
+\begin{center}
+\includegraphics[scale=0.45]{approaches.pdf}
+\end{center}
+\vspace{-20px}
+\caption{\small (a) With the language-external approach, novel constructs are packaged into separate languages.
+Users can only safely and naturally call into languages if the interface uses common constructs and an interoperability layer has been developed (the \emph{interoperability problem}).
+(b) With the language-internal approach, there is one extensible host language and the compile-time and edit-time logic governing novel constructs is expressed within libraries.
+If the extension mechanism guarantees the safety of arbitrary compositions of extensions, the necessary primitives can simply be imported by library clients as needed, and so interoperability is not a problem.}
+\label{approaches}
+\end{figure}
+Designing and implementing a programming language together with its supporting tools (collectively, a \emph{programming system}) that has a sound theoretical foundation, helps users identify and fix errors as early as possible, supports a natural programming style, and that performs well across diverse problem domains and hardware platforms remains a grand challenge in computing.
+No small, fixed collection of primitive abstractions and tools has been found to fully satisfy these criteria in all situations.
+Rather, researchers and domain experts  continue to design and implement novel abstractions, provide alternative implementations of existing abstractions, and supplement languages with new tools in order to better satisfy these criteria within the constraints of different problem domains.
 
-Designing and implementing a programming language and its supporting tools (collectively, a \emph{programming system}) that simultaneously supports a natural programming style, helps users find and fix errors early, and performs well across diverse problem domains remains a grand challenge in computing.
-Indeed, it is now widely acknowledged that no simple set of constructs will be able to handle all of the challenges modern computing faces.
-Consistent with this view, a large number of researchers and domain experts in both research and industry are actively engaged in the design, implementation and analysis of a variety of novel language-based abstractions, compilation strategies and tools targeted at different problem domains.
+To realize a new abstraction or system behavior, such experts can choose either a \emph{language-internal approach}, where they work within an existing language and distribute their solutions as libraries, or a \emph{language-external approach}, where they either derive a new programming system or extend an existing system by some mechanism that is not part of the language itself, such as an extension mechanism supported by a {particular} compiler, editor or other tool.
 
-%In implementing a new feature or abstraction, these experts can take one of several approaches. 
-To realize a new feature or abstraction, these experts often take a \emph{language-oriented approach}, whereby the new constructs are developed and distributed together with a new programming language.
-The flexibility of this approach, coupled with support from tools like compiler generators, has led to a proliferation of specialized languages, each built around a particular set of privileged constructs.
-Unfortunately, a key issue limiting the practicality of this approach occurs at the boundaries between languages -- specialized constructs particular to one language cannot always be safely and naturally expressed in another, so building a program out of components written in different specialized languages can often prove difficult.
-
-One common strategy that partially addresses this problem is to target an established language and runtime, such as the Java Virtual Machine (JVM), and support a superset of its features.
-This only solves the interoperability problem in one direction.
-While calling from the new language to the old language becomes straightforward, calls in the other direction, or between languages sharing a common target, remain restricted by the constructs available in the common language.
-This issue is illustrated in Figure \ref{approaches}a.
-This restriction can  constrain the design of new languages to only include constructs that can be expressed reasonably naturally in the target language.
-More novel innovations are often difficult to translate into familiar forms, however.
-For instance, the module system in F\# is severely constrained relative to the module system in its predecessor, OCaml, due to the need for interoperability with other CLI languages.
-
-Stepping back, it is clear that a \emph{library-oriented approach}, shown in Figure \ref{approaches}b, where new constructs are implemented entirely within a single host language as libraries, should be preferred when possible.
+When possible, taking the \emph{language-internal approach} is significantly simpler and more practical.
+If an abstraction or system behavior can be realized by creatively repurposing existing language constructs, then  providers and clients face fewer barriers to adoption.
 % -- library-based constructs cannot interfere with one another (assuming a suitable namespacing standard), libraries benefit from common distribution infrastructure and tool support, and externally-accessible library interfaces need not be restricted to a fixed common subset of all available constructs (e.g.
@@
 We diagram this key latter distinction in Figure \ref{approaches}.
-Of course, this is very often \emph{not} possible today because from the perspective of a library developer, the programming system is a largely inflexible, \emph{monolithic} entity.
-That is, the language's syntax, type checking rules and semantics are fixed in advance, the compiler is a black box implementation of these rules, and the tools operate according to domain-agnostic protocols that use only the declared structure of the program being developed as input.
-To enhance any of these aspects of the programming system, the expert must step out of the language, and often out of the programming system entirely.
+Unfortunately, this is very often \emph{not} possible today because from the perspective of library code, the programming system is a largely inflexible, \emph{monolithic} entity.
+That is, the language's syntax and its static and dynamic semantics are fixed in advance, the compiler is a ``black box'' implementation of these fixed semantics, and other tools operate according to domain-agnostic protocols that use only the basic structure of the program as input.
+To supplement any of these components of the programming system, experts must step out of the language, and often out of the programming system entirely.
 
-To put it another way, monolithic programming languages enforce a fundamental dichotomy between \emph{first-class constructs}, which enjoy support throughout the programming system but must be designed and implemented by the system designer in advance, and \emph{user-defined constructs}, which can be distributed as user libraries, but must creatively combine and repurpose the small set of available first-class constructs to express all desired run-time, compile-time and edit-time behaviors.
-This dichotomy has forced researchers and domain experts who want to significantly advance the state-of-the-art in programming systems, particularly in compiled, statically-typed systems where this dichotomy is most salient, to design a new language in many cases.
-But this then couples their core innovations with a collection of other, unrelated design choices, requires more effort both for the experts developing the innovation and its targetted user community, and leads to the intrinsic interoperability problems described above for users developing applications consisting of components written in different languages.
+In these situations, experts must take a \emph{language-external approach}, often by developing and distributing new constructs together with a new language and developing an array of necessary tools.
+This increases the development burden on these experts, although tools like compiler generators and language workbenches have started to simplify this process, and also requires experts to couple their innovations with a collection of other unrelated design choices, making adoption more difficult for clients.
 
-\begin{figure}
-\begin{center}
-\includegraphics[scale=0.5]{approaches.pdf}
-\end{center}
-\vspace{-20px}
-\caption{\small (a) With the language-oriented approach, novel constructs are packaged into separate languages.
-Users can only safely and naturally call into languages consisting of common constructs.
-(b) With the library-oriented approach, there is one language and novel constructs are packaged as normal libraries.
-Thus, interoperability is not a problem but in monolithic systems, this approach is less expressive.\label{approaches}}
-\end{figure}
+A more fundamental limitation of the language-external approach arises at the interface between languages.
+The specialized constructs particular to one language cannot always be safely and naturally expressed in another, so building a program out of components written in different languages, when such constructs are used at the interface boundaries, can be difficult.
+We refer to this as the \emph{interoperability problem}.
+One increasingly common strategy taken by language designers to partially address the interoperability problem is to target an established language and runtime, such as the Java Virtual Machine (JVM), and support a superset of its features.
+Scala \cite{scala} and F\# \cite{fsharp} are examples of languages that have taken this approach.
+This addresses interoperability in only one direction (Figure \ref{approaches}a).
+While calling into the common language becomes straightforward, calls in the other direction, or between languages sharing the common target, are still restricted by the constructs available in the common language.
+If interoperability in both directions is desired, the new languages must only include constructs that can be expressed safely and naturally in the target language.
+More novel innovations, however, are often difficult to define in terms of familiar forms in ways that guarantee all necessary invariants are maintained and that are reasonably natural for users of other languages.
+In F\#, for example, the type system guarantees that null values cannot occur, but this invariant is not maintained when calling from other languages.
+The F\# type system also includes support for units of measure \cite{fsharpunits}, but such domain-specific invariants cannot be guaranteed when calling into F\# from other CLI languages.
+Similarly, interfaces built using Scala traits can be difficult to implement from Java or other JVM languages.
+In some cases, features are omitted entirely due to this requirement.
+For example, the module system in F\# is significantly limited relative to the module system in its predecessor, OCaml, due to the need for bidirectional interoperability between other languages on the common language infrastructure (CLI).
+
+For these reasons, we argue that language-external approaches should be considered harmful.
+The goal of the research proposed here is to fundamentally reorganize the core components of the programming system so that such language-external approaches are less frequently necessary, by designing \emph{language-internal} mechanisms that give developers control, from within the language itself, over edit-time and compile-time behaviors that are currently centrally controlled by language and tool designers and design committees\footnote{To be a bit piquant, one might compare today's monolithic programming languages to isolationist {centrally-planned} economies, whereas extensible languages more closely resemble modern market economies.
+We leave fleshing out such analogies to the reader.}.
+In particular, we will describe how parsing, typechecking, translation and code completion can all be influenced by user-defined constructs \todo{describe what this looks like - libraries + reference Fig 1b.} in ways that preserve the safety of the language as a whole and admit strong composition principles \todo{elaborate on safety requirements, merge with next paragraph}.
+
+The extension mechanism must be expressive enough to allow users to associate rich run-time, compile-time and edit-time behaviors with user constructs directly, while being sufficiently restrictive to maintain the global safety properties of the language and system as a whole, and to ensure that constructs cannot interfere with one another.
+This explicit consideration of the full programming system, and the tension between expressiveness and safety, are key aspects of our approach.

### 2013-09-10 thesis-proposal/acc.tex
updated proposal

--- 2013-08-26 thesis-proposal/conclusion.tex
@@
 % !TEX root = omar-thesis-proposal.tex
-\section{Conclusion}
-
-Monolithic programming systems enforce a dichotomy between \emph{built-in constructs}, which enjoy support throughout the programming system but must be designed and implemented by the system designer in advance, and \emph{user-defined constructs}, which can be distributed as user libraries, but must creatively combine and repurpose the small set of available built-in constructs to express all desired run-time, compile-time and edit-time behaviors.
-This dichotomy has forced researchers and domain experts who want to significantly advance the state-of-the-art in programming systems, particularly in compiled, statically-typed systems where this dichotomy is most salient, to design and develop new languages.
-But this then couples their core innovations with a collection of unrelated design decisions, requires more effort both for the experts developing the innovation and its targeted user community, and leads to intrinsic problems for users developing applications consisting of components written in different languages.
-\newpage
-\listoftodos
-\newpage
+\section{Active Code Completion}

### 2013-09-10 thesis-proposal/problem.tex
updated proposal

@@
 % !TEX root = omar-thesis-proposal.tex
-\section{Introduction}\begin{figure}
+\section{Motivation}\label{motivation}\begin{figure}
 \begin{center}
@@
 Designing and implementing a programming language together with its supporting tools (collectively, a \emph{programming system}) that has a sound theoretical foundation, helps users identify and fix errors as early as possible, supports a natural programming style, and that performs well across diverse problem domains and hardware platforms remains a grand challenge in computing.
-No small, fixed collection of primitive abstractions and tools has been found to fully satisfy these criteria in all situations.
-Rather, researchers and domain experts  continue to design and implement novel abstractions, provide alternative implementations of existing abstractions, and supplement languages with new tools in order to better satisfy these criteria within the constraints of different problem domains.
+It has become clear that no small, fixed collection of primitive abstractions and tools can fully satisfy these criteria in all situations.
+Rather, researchers and domain experts must continue, in a principled manner, to design and implement novel abstractions, provide alternative implementations of existing abstractions, and supplement languages with new tools, and tools with new behaviors, to better satisfy these criteria within the constraints of different problem domains.
 
-To realize a new abstraction or system behavior, such experts can choose either a \emph{language-internal approach}, where they work within an existing language and distribute their solutions as libraries, or a \emph{language-external approach}, where they either derive a new programming system or extend an existing system by some mechanism that is not part of the language itself, such as an extension mechanism supported by a {particular} compiler, editor or other tool.
+%To realize a new abstraction or system behavior, such experts can consider either a \emph{language-internal approach}, where they work within an existing language and distribute their solutions as libraries, or a \emph{language-external approach}, where they create a new, distinct programming system (often centered around what has come to be called a new \emph{domain-specific language} \cite{dsl}) or extend an existing system by some mechanism that is not part of the language itself, such as an extension mechanism supported by a {particular} compiler, editor or other tool.
 
-When possible, taking the \emph{language-internal approach} is significantly simpler and more practical.
-If an abstraction or system behavior can be realized by creatively repurposing existing language constructs, then  providers and clients face fewer barriers to adoption.
-% -- library-based constructs cannot interfere with one another (assuming a suitable namespacing standard), libraries benefit from common distribution infrastructure and tool support, and externally-accessible library interfaces need not be restricted to a fixed common subset of all available constructs (e.g.
-the object system of the JVM or CLI).
-We diagram this key latter distinction in Figure \ref{approaches}.
-Unfortunately, this is very often \emph{not} possible today because from the perspective of library code, the programming system is a largely inflexible, \emph{monolithic} entity.
-That is, the language's syntax and its static and dynamic semantics are fixed in advance, the compiler is a ``black box'' implementation of these fixed semantics, and other tools operate according to domain-agnostic protocols that use only the basic structure of the program as input.
-To supplement any of these components of the programming system, experts must step out of the language, and often out of the programming system entirely.
+When possible, taking a \emph{language-internal approach} to implementing a new abstraction or system behavior (collectively, a new \emph{feature}) is simpler and more practical.
+If a feature can be realized by creatively repurposing existing features and distributed as a library, both providers and clients face fewer barriers to adoption because it is easy to gradually integrate library-based features into existing code and they leverage well-understood and well-developed mechanisms.
+But taking this approach is often \emph{not} possible today because libraries are generally vehicles for specifying the run-time behavior of a program.
+The compile-time and edit-time behaviors of a programming system cannot easily be adapted or extended in a specialized way from within libraries in most programming systems.
+That is, the language's syntax, static semantics and corresponding dynamic semantics are fixed in advance, the compiler is a ``black box'' implementation of these fixed semantics, and the other tools, like code editors and debuggers, operate according to fixed, domain-agnostic protocols based on these fixed semantics.
 
-In these situations, experts must take a \emph{language-external approach}, often by developing and distributing new constructs together with a new language and developing an array of necessary tools.
-This increases the development burden on these experts, although tools like compiler generators and language workbenches have started to simplify this process, and also requires experts to couple their innovations with a collection of other unrelated design choices, making adoption more difficult for clients.
+If any of these components of the system must be modified to realize a new feature, providers must instead take a \emph{language-external approach}, either by developing a new or derivative programming system (often centered around a so-called \emph{domain-specific language} \cite{dsl}) or by extending an existing system by some mechanism that is not part of the language itself, such as an extension mechanism supported by a {particular} compiler or other tool.
+%Extending a particular tool is morally equivalent, for our purposes, to creating a new system, because we aim to minimize the extent to which the edit-time, compile-time and run-time semantics of libraries depend on the implementation details of a particular tools.
 
-A more fundamental limitation of the language-external approach arises at the interface between languages.
-The specialized constructs particular to one language cannot always be safely and naturally expressed in another, so building a program out of components written in different languages, when such constructs are used at the interface boundaries, can be difficult.
+Unfortunately, when providers of new features take language-external approaches, it causes problems for clients.
+Features become coupled to a collection of other unrelated and often underdeveloped features, making adoption more costly when these other choices are not appropriate.
+For example, although recent evidence suggests that developers prefer language-integrated parallel programming abstractions to library-based implementations if all else is equal \cite{langvslib}, library-based implementations remain more widely adopted because they do not require porting applications to a language designed around a few privileged parallel programming abstractions.
+These abstractions may only be useful in performance-critical portions of the application, and then only in cases where no competing abstraction would be more appropriate.
+
+%\item Developing a new language and its associated tools places a significant development burden on providers who may wish only to promote a few core innovations, although tools like compiler generators, language workbenches and easy-to-extend tools can decrease this burden.
+%\item
+
+%Clients seem to prioritize the ability to choose different features for different portions of an application. 
+If it were easy to call between languages, then having a variety of specialized languages would be less of a problem.
+In fact, some have advocated for just this development model, calling it the the \emph{language-oriented approach} to software development \cite{languageoriented}.
+A fundamental problem with this approach, however, arises at the interfaces between languages.
+The specialized abstractions particular to one language cannot be safely and naturally expressed in another, in general, so building a program out of components written in many different languages is difficult or impossible whenever these specialized abstractions are exposed at library interface boundaries.
 We refer to this as the \emph{interoperability problem}.
-One increasingly common strategy taken by language designers to partially address the interoperability problem is to target an established language and runtime, such as the Java Virtual Machine (JVM), and support a superset of its features.
-Scala \cite{scala} and F\# \cite{fsharp} are examples of languages that have taken this approach.
-This addresses interoperability in only one direction (Figure \ref{approaches}a).
-While calling into the common language becomes straightforward, calls in the other direction, or between languages sharing the common target, are still restricted by the constructs available in the common language.
-If interoperability in both directions is desired, the new languages must only include constructs that can be expressed safely and naturally in the target language.
-More novel innovations, however, are often difficult to define in terms of familiar forms in ways that guarantee all necessary invariants are maintained and that are reasonably natural for users of other languages.
-In F\#, for example, the type system guarantees that null values cannot occur, but this invariant is not maintained when calling from other languages.
-The F\# type system also includes support for units of measure \cite{fsharpunits}, but such domain-specific invariants cannot be guaranteed when calling into F\# from other CLI languages.
-Similarly, interfaces built using Scala traits can be difficult to implement from Java or other JVM languages.
-In some cases, features are omitted entirely due to this requirement.
-For example, the module system in F\# is significantly limited relative to the module system in its predecessor, OCaml, due to the need for bidirectional interoperability between other languages on the common language infrastructure (CLI).
 
-For these reasons, we argue that language-external approaches should be considered harmful.
-The goal of the research proposed here is to fundamentally reorganize the core components of the programming system so that such language-external approaches are less frequently necessary, by designing \emph{language-internal} mechanisms that give developers control, from within the language itself, over edit-time and compile-time behaviors that are currently centrally controlled by language and tool designers and design committees\footnote{To be a bit piquant, one might compare today's monolithic programming languages to isolationist {centrally-planned} economies, whereas extensible languages more closely resemble modern market economies.
-We leave fleshing out such analogies to the reader.}.
-In particular, we will describe how parsing, typechecking, translation and code completion can all be influenced by user-defined constructs \todo{describe what this looks like - libraries + reference Fig 1b.} in ways that preserve the safety of the language as a whole and admit strong composition principles \todo{elaborate on safety requirements, merge with next paragraph}.
+One strategy increasingly being taken by programming language designers to partially address the interoperability problem today is to target an established language, such as the Java Virtual Machine (JVM) bytecode, and support a superset of its constructs.
+Scala \cite{scala} and F\# \cite{fsharp} are examples of general-purpose languages that have taken this approach.
+This only provides full interoperability in one direction (Figure \ref{approaches}a).
+While calling into the common language becomes straightforward, calls in the other direction, or between the languages sharing the common target, are still restricted by the constructs available in the common language.
+If full interoperability is desired, new languages must only include constructs that can already be expressed safely and reasonably naturally in the common language.
+Many innovative features, however, can be difficult to define in terms of existing features in ways that guarantee all necessary invariants are maintained and that do not require large amounts of boilerplate code.
+For example, the type system of F\# guarantees that null values cannot occur within F\# data structures, but maintaining this important invariant still requires run-time checks because the typing rules of F\# do not apply when F\# code is called from other languages on the Common Language Infrastructure (CLI) like C\#.
+The F\# type system also includes support for checking that units of measure are used correctly \cite{fsharpunits}, but this domain-specific invariant is left completely unchecked at language boundaries.
+In Scala, interfaces built around traits that have default method implementations are difficult to implement from Java or other JVM languages and the workaround can break if the trait is modified \cite{scalatraitinterop}.
+In some cases, desirable features must be omitted entirely due to concerns about  interoperability.
+The module system in F\#, for example, is substantially simpler than that in its predecessor, OCaml, despite F\# otherwise aiming to maintain compatibility, due to the need for bidirectional interoperability between F\# and other CLI languages.
+%\end{itemize}
 
-The extension mechanism must be expressive enough to allow users to associate rich run-time, compile-time and edit-time behaviors with user constructs directly, while being sufficiently restrictive to maintain the global safety properties of the language and system as a whole, and to ensure that constructs cannot interfere with one another.
-This explicit consideration of the full programming system, and the tension between expressiveness and safety, are key aspects of our approach.
+For these reasons, we argue that the language-oriented approach is fundamentally flawed and so taking a language-external approach to realizing a new feature, where new features are packaged into mutually incompatible languages and associated tools, should be considered harmful and avoided when possible.
+The goal of the research proposed here is to fundamentally reorganize the core components of the programming system so that such language-external approaches are less frequently necessary, by designing \emph{language-internal extension mechanisms} that give developers control over edit-time and compile-time behaviors that have previously been centrally-controlled by  the  committees that govern languages or tools\footnote{One might compare today's monolithic programming systems to  {centrally-planned} economies, whereas extensible\- languages more closely resemble modern market economies.}\todo{read Arch D.
+Robison.
+Impact of economics on compiler optimization.}.
+Specifically, we will show how control over \textbf{parsing}, \textbf{typechecking}, \textbf{translation} (the first stage of compilation) and \textbf{code completion} can be delegated to user-defined logic distributed in {libraries}, as illustrated in Figure \ref{approaches}b.
+Such libraries are called \emph{active libraries} because, rather than being passive consumers of features already available in the system, they contain logic invoked directly by tools to provide new domain-specific functionality during development or compilation \cite{active-libraries}.
+Features implemented within active libraries can be imported as needed by clients of libraries that rely on them, unlike features implemented by language-external means.
+
+Some critical issues having to do with {safety} must be overcome before library-based extension mechanisms can be introduced into a programming system, because if too much control over such core aspects of the system is given to users, the system may become unreliable.
+For example, an extension could weaken important metatheoretic guarantees previously provided by the system.
+Type safety, for instance, may no longer hold if the static and dynamic semantics of the language can be modified arbitrarily within libraries.
+Furthermore\-, even if extensions can be shown not to cause such problems in isolation, there may still be conflicts between extensions that could lead to subtle problems at link-time.
+For example, if two active libraries defined differing semantics for the same syntactic form, the issue would only manifest itself when both libraries were imported somewhere within the same program.
+These kinds of safety issues have plagued previous attempts at language-internal extensibility (discussed further in the next section).
+To prevent them, the system must strictly constrain extensions such that they are guaranteed safe in isolation and also safely composable in any combination.
+
+The mechanisms described in this thesis are designed to be highly \textbf{expressive}, allowing library-based implementations of features comparable to built-in features found in modern programming systems, without compromising on these important notions of \textbf{safety}.
+This is accomplished by organizing extension logic around types, rather than languages.
+By scoping new compile-time and edit-time logic to expressions of the type with which it is associated, rather than applying it globally or only within contiguous syntactic blocks as in many previously-developed mechanisms, conflicts are avoided.
+We call types with such logic associated with them \emph{active types} and systems that support them \emph{actively-typed programming systems}.
+By constraining the extension logic with a type system and using techniques from the compiler correctness literature, we can ensure that the system as a whole maintains important safety properties.
+%This represents a minimalist approach to system design -- the conventional distinction between built-in and user-defined constructs is blurred and most features of the system are orthogonally implemented as {libraries}, rather than by the maintainers of the system.
+
+%The mechanisms we describe will do so primarily by delimiting the scope of an extension to expressions of a single user-defined type or family of types.
+
+%This can be thought of as a more pernicious form of the conflict that arises when two globally-accessible constructs are given the same name.
+n languages without universal namespacing mechanisms (e.g.
+C, JavaScript, \LaTeX, ML and many others).
+
+%The extension mechanism\todo{elaborate on safety requirements + tension between expressiveness and safety, merge with next paragraph}.
+must be expressive enough to allow users to associate rich run-time, compile-time and edit-time behaviors with user constructs directly, while being sufficiently restrictive to maintain the global safety properties of the language and system as a whole, and to ensure that constructs cannot interfere with one another.

### 2013-09-10 thesis-proposal/work.tex
updated proposal

--- 2013-01-09 thesis-proposal/work.tex
@@
 This will allow us to understand the challenges and opportunities of using a dependently-typed functional language at the core of an extensible programming system.
-
-\subsection{Active Code Prediction}
-\subsubsection{Structured, Statistical Code Prediction}
-
-Programming languages are formal systems with rich syntactic and semantic structure. They are
-also human systems, in that they are used extensively by people in patterned ways to express their
-intent. Many tools are designed to help people write code more efficiently by predicting
-the source code that a developer intends.
-For example, code completion systems for editors like Eclipse for Java display pop-up menus containing the members of the class of the variable being manipulated by the developer.
-
-Typical code completion systems use only this sort of semantic
-information about the language and the libraries being used, and do not incorporate data about how developers have written programs
-in the past. Recent work by Hindle et al.~\cite{Hindle:2012:NS:2337223.2337322} demonstrated, however, that source code could be successfully predicted statistically
-using a simple $n$-gram model that used a tokenized, rather than structural, representation of source code, trained on existing codebases.
-
-Our first aim is to unite the structured and statistical approaches to source code prediction.
-That is, rather than using a tokenized representation of source code, we would like to do statistical prediction on a more natural representation of source code: the typed syntax tree.
-We can then condition our predictions using semantic information, specifically:
-\begin{itemize}
-\item the \emph{type}, denoted $\tau$, of the expression being predicted (e.g. \verb|int| or \verb|Color|)
-\item the \emph{syntactic context}, denoted $\sigma$, in which the expression occurs (e.g. whether the expression is an argument of a function call, the guard of an \verb|if| statement, etc.)
-\item the \emph{program context}, denoted $\Gamma$, in which the expression occurs (i.e. the set of variables paired with their types that are in scope at the location that the expression occurs.)
-\end{itemize}
-
-For example, if a user has entered the code \texttt{Planet destination = }, where \verb|Planet| is an enumeration containing \verb|Mercury|, \verb|Venus|, \verb|Earth|, etc.
-(but not \verb|Pluto|, of course), then we have that the expected type at the cursor is \verb|Planet|, the syntactic context is \emph{assignment}, and given a program context, our prediction space should only assign non-zero probabilities to:
-\begin{itemize}
-\item literal members of the \verb|Planet| enumeration (e.g. \verb|Planet.EARTH|)
-\item variables and fields of type \verb|Planet| available from the program context
-\item calls to methods available from the program context that have return type \texttt{Planet}\footnote{We can consider  operators like $+$ and $[]$ as methods of the built-in types in Java.}
-\end{itemize}
-To assign a particular probability to an expression, denoted $e$, we first determine how likely it is that the expression is of each of these three \emph{syntactic forms}, where syntactic forms are denoted $\phi$.
-For each form, we can then assign probabilities to particular expressions of that form according to some form-specific conditional distribution.
-This model can be understood as a graphical model, as shown in Figure \ref{graphicalmodel}, where the syntactic form is a latent variable.
-The conditional distributions for both the syntactic form and expression are learned using data gathered from analyses of prior code corpuses (smoothed using a suitable method in cases where enough information is not available).
-\begin{figure}
-    \begin{displaymath}
-    \xymatrix{
-      *+[F]{\color{ForestGreen}\txt{Type ($\tau$)}} \ar[dr] \ar[drr] & & \\
-      & *+[F]{\color{black} \txt{Syntactic\\Form ($\phi$)}} \ar[r] & *+[F]{\color{cyan}\txt{Expression ($e$)}}  \\
-      *+[F]{\color{ForestGreen}\txt{Syntactic\\Context ($\sigma$)}} \ar[ur] \ar[urr] & & *+[F]{\color{ForestGreen}\txt{Program\\Context ($\Gamma$)}} \ar[u]  
-    }
-    \end{displaymath}
-\caption{\small A graphical model representing our approach.
-Green variables are always observed (we do not assign marginal distributions to them.) The syntactic form, $\phi$, is a latent variable, and the expression, $e$, is unknown.
-Note that the syntactic form is a function of the expression.
-\label{graphicalmodel}}
-\end{figure}
-
-\subsubsection{Active Code Prediction}
-Given a system for predicting expressions in the context of a type syntax tree, we can then naturally ask: what about specialized domain knowledge that cannot be readily extracted from prior code analysis?
-For example, consider a type representing strings in a particular language, such as English.
-Constructors of the \verb|EnglishString| type should be biased to accept literals that have high likelihoods according to some natural language model, but we do not wish to encode all possible natural language models in the core of our prediction system.
-We can instead take an active types approach by allowing users to equip a type, such as \verb|EnglishString|, with a function, $f_{\text{prob}}$, that takes on the job of predicting the probabilities of only expressions of that type.
-That is, given a reified representation of $\sigma$, $\Gamma$ and $\phi$, it should be able to produce $P(e | \tau, \phi, \sigma, \Gamma)$.
-We can then use this prediction directly as if it had been produced by a method described above.
-
-\subsubsection{Foundations, Implementation and Analysis}
-We propose developing the theoretical foundations of this idea, in terms of probability theory and type theory, equipping it with a notion of active code prediction as described above, and implementing this atop a real language, likely Java.
-We have sketched and prototyped a small version of this as an Eclipse extension in the context of a prior course project.
-We will then evaluate the accuracy of these models, develop use cases as in the related work on active code completion described in Section \ref{graphite}, and analyze the usability and usefulness of this prediction system in applied contexts.
-In particular, we are interested in uses within code editors to predict expressions and portions of expressions, as well as uses within programming systems for disabled and paralyzed individuals, who need predictive text entry to be able to communicate at a reasonable pace.
-The information theoretic foundations for this latter application, in the context of brain-computer interfaces, have been developed as prior work by one of the investigators on this proposal \cite{journals/ijhci/OmarAJBMMMC11}.
-
-\subsection{Syntax and Visual Representation}
-Thus far, we have been discussing active types within the context of languages with a fixed syntax.
-However, in many specialized domains, a specialized syntax or visual representation for novel constructs may be appropriate.
-There are several approaches that we can take to include extensible syntax support in an actively-typed programming system.
-We have not settled on any one of these, so our proposed research will include a rigorous examination of the pros and cons of each approach.
-
-\subsubsection{Extensible Parsing} 
-Extensible parsers have been well-studied in the literature (e.g.
-\cite{journals/entcs/BrabrandSV03}), and such systems could be integrated with an actively-typed language without conflict with one caveat: each new syntactic form needs to have a dispatch protocol associated with it, to assign control over its semantics to one of its subexpressions.
-Such a mechanism will be one avenue we will evaluate for suitability.
-
-\subsubsection{Flexible Reinterpretation}
-An approach that we have prototyped, but not extensively utilized, in Ace is to allow active types to directly observe the syntax trees of subexpressions, rather than simply their types as we have described above.
-This allows a highly contextualized reinterpretation of a syntax tree, similar to capabilities found in macro systems (although without a direct rewriting component.) We will further explore this approach, particularly as we begin implementing abstractions that may not naturally fit within Python's existing syntactic forms in an intuitive way.
-
-\subsubsection{Delimited Parsing}
-A related alternative approach would be to partially parse a file, leaving delimited regions either unparsed or only tokenized and dispatching to an active type to do the actual parsing.
-This would give more flexibility to the active type, but also a greater responsibility to fully parse and give semantics to the provided source code.
-Nevertheless, this may be appropriate in certain cases where a highly-specialized local syntax is desirable for convenience (and flexible reinterpretation or active code completion at edit-time does not suffice.)
-
-\subsubsection{Actively-Typed Structured Editors}
-Projects such as Citrus \cite{Ko05} and Barista \cite{Ko06} are analogs of language frameworks for structured editors -- editors where source code elements are represented directly as a tree in a visual or semi-visual form, rather than written as a string as in conventional languages.
-These tools demonstrate that modular construction of such editors may be possible.
-An actively-typed structured editor would instead allow an active type to directly control the visual representation and interaction behaviors associated with expressions of that type.
-This is the most direct analog of active types in this domain, but requires a somewhat radical shift in program representation.
-It is unlikely that this can be integrated easily into existing languages and systems, as we have largely done in our other work.
-Thus, we will develop a separate prototype of this approach, and believe that the long-term evolution of actively-typed programming systems will be toward such a representation.

### 2013-09-17 thesis-proposal/background.tex
Updated abstract, most of introduction, beginning of approach
updated proposal
Updated beginning of proposal

@@
-\section{Background}
-We review existing approaches that are available to researchers and domain experts who wish to develop novel constructs below.
+% !TEX root = omar-thesis-proposal.tex
+\subsubsection{Background: Active Libraries}\label{alibs}
+The term \emph{active libraries} was first used by Veldhuizen et al.
+\cite{atthesis, atpaper} to describe ``libraries that take an active role in compilation, rather than being passive collections of subroutines''.
+The authors went on to suggest a number of concrete reasons libraries might benefit by being able to influence the programming system at compile-time or edit-time, including high-level program optimization, checking programs for correctness against specialized criteria, reporting domain-specific errors and warnings, and ``rendering domain-specific textual and non-textual program representations and for interacting with such representations'' (anticipating interactions between libraries and tools other than just the compiler.
+Our definition, given in Section \ref{motivation}, differs  from theirs, partly for this reason).
 
-\subsection{Language-Oriented Approaches}
-\subsubsection{Language Frameworks}
-A number of tools have been developed to assist with this task of developing new languages and their associated tools, including compiler generators, language workbenches and domain-specific language frameworks (see \cite{fowler2010domain} for a review).
+% This stuff isn't actually hooking into the compiler...
+The first concrete realizations of active libraries, prompting the introduction of the term, were libraries that performed domain-specific program optimization at compile-time by exploiting language mechanisms that allow for limited compile-time computation.
+A prominent example in the literature is Blitz++, a library that uses C++ template metaprogramming to optimize compound operations on vectors and matrices by eliminating intermediate allocations \cite{blitzpp}.
+Although this and several other interesting optimizations have been shown possible by this technique, its expressiveness is fundamentally limited because template expansion allow for only the substitution of compile-time constants into pre-written code, and template metaprograms are notoriously difficult to read, write, debug and reason about.
+%They are, however, composable, because templates are associated with a single function or class, and  the only metatheoretic issue is that template expansion can be non-terminating.
+
+More powerful compile-time \emph{term rewriting mechanisms} integrated into some languages can also be used for optimization, as well as for inserting specialized error checking and reporting\- logic and extending the language with powerful domain-specific abstractions.
+These mechanisms are highly expressive because they allow users to programmatically manipulate syntax trees directly, but they suffer from problems of composability and safety.
+Macros, such as those in MetaML \cite{metaml}, Template Haskell \cite{template_haskell} and others\todo{reference other macro systems}, can take full control over all the code that they enclose, so there is no way to guarantee that nested macros will compose as intended.
+Outer macros can neglect to invoke or  manipulate the output of inner macros in ways that can cause conflicts or weaken important guarantees.
+Once data escapes a macro's scope, there is no way to rely on the guarantees and features that were available within its scope -- the output of a macro is a value in the underlying language, so the underlying language is all that governs it (a problem related to the interoperability problem of Section \ref{motivation}).
+It can also be difficult to reason about the semantics of code when a number of enclosing macros may be manipulating it.
+
+Some rewriting systems go beyond the block scoping of macros.
+Xroma (pronounced ``Chroma''), for example, is designed around active libraries and allows users to insert custom rewriting and error checking passes into the compiler from within libraries \cite{xroma}.
+Similarly, the derivative of Haskell implemented by the Glasgow Haskell Compiler (GHC) allows libraries to define custom compile-time term rewriting logic if an appropriate flag is passed in \cite{haskell-rewrite}.
+In both cases, the user-defined logic can dispatch on arbitrary patterns of code throughout the component or program the extension is active in, so these mechanisms are highly expressive and avoid some of the difficulties of block-scoped macros.
+But libraries containing such global rewriting logic cannot be safely composed because two different libraries may attempt to rewrite the same piece of code differently.
+It is also difficult to guarantee that such logic is correct and difficult for users to reason about code when simply importing a library can change the semantics of the program in a non-local manner.
+% http://www.haskell.org/ghc/docs/latest/html/users_guide/rewrite-rules.html
+
+Another example of an active library approach to extensibility is SugarJ \cite{sugarj}.
+SugarJ allows libraries to extend the base syntax of Java in a nearly-arbitrary manner, and these extensions are imported transitively throughout a program.
+SugarJ libraries are thus also not safely composable.
+For example, a library that defines a literal syntax for HTML would conflict with another that defines a literal syntax for XML because they define differing semantics for some of the same syntactic forms.
+If SugarJ was used by two different regular expression engines to provide literal syntax for regular expression patterns, there could easily be conflicts at link-time because both will introduce many of the same standard notations but back them with differing implementations.
+And again, it can be difficult for users to predict what type of data an unfamiliar piece of syntax desugars into, leading to difficulties reading and reasoning about code.
+
+%Another example comes from the Tau package for tuning and analyzing parallel programs, which allows libraries to instrument themselves declaratively so that internal details can be hidden from users of the tool \cite{tau}.
+
+%The researchers behind these tools often claim that such conflicts are rare.
+There have, to our knowledge, been no studies of this claim.
+Nevertheless, the core fact remains: conflicts can arise, which makes it difficult to rely on such libraries
+
+%There also does not appear to be a clear, well-founded theoretical foundation for this approach, nor for active libraries in general.
+
+%\todo{Smooth this transition}Developers specify type-specific compile-time and edit-time features by equipping type families, when they are defined, with functions written in an appropriate metalanguage.
+These are selectively invoked by tools, such as the parser, type checker, translator and editor, when they work with expressions of that type, and only in those situations.
+
+%\subsection{Example: First-Class Natural Numbers}
+%To make our approach more concrete and emphasize its foundational character, let us begin with a very simple example, first-class natural numbers.
+If a programming system included first-class support for natural numbers, it might support behaviors like the following:
+%
+%\begin{enumerate}
+%\item type checking rules that constrain uses of the three operators associated with the $\nat$ type, as defined in G\"{o}del's System \textbf{T} \cite{tapl}: $\natz$, $\nats{e}$ and $\natrec{e_1}{e_2}{e_3}$\todo{phrasing of his}
+%\item type checker error messages that provide domain-specific assistance, like detecting when the trailing $\textbf{z}$ in a constant has been omitted: \todo{phrasing / merge this into 1?}
+%\begin{verbatim}
+%    s(s(s))
+%        ^ (Line 1, Character 5)
+%    Error: Operator s requires 1 argument of type nat. Provided 0 arguments.
+%    Did you mean s(z)?
+%\end{verbatim}
+%\item translation into an efficient integer representation at run-time
+%\item syntax support \todo{elaborate on syntax support for nat literals}
+%\item editor support, such as the ability to instantiate nat constants more easily \todo{elaborate on this / reference Graphite section}
+%%generate statistical predictions useful for code completion (e.g. biasing smaller constants) and other editor interactions, customized syntactic/visual representations and so on.
+%\end{enumerate}
+%
+
+%\section{Related Work}
+%We review existing approaches that are available to researchers and domain experts who wish to develop novel constructs below.
+%
+%\subsection{Language-Oriented Approaches}
+%\subsubsection{Language Frameworks}
+%A number of tools have been developed to assist with this task of developing new languages and their associated tools, including compiler generators, language workbenches and domain-specific language frameworks (see \cite{fowler2010domain} for a review).
 In some cases, these tools allow language features to be defined modularly and composed differentially to produce a variety of different languages.
@@
 User-defined code is still ultimately compiled against a particular language, and because language composition cannot be automatic or guaranteed safe, interoperability with code written in other languages is thus limited by the difficulties described above despite the modular construction.
-
-\subsubsection{Extensible Compilers and Tools}
-A related methodology is to implement language features as compiler and tool extensions directly.
+%
+%\subsubsection{Extensible Compilers and Tools}
+%A related methodology is to implement language features as compiler and tool extensions directly.
 A number of extensible compilers have been developed to support this approach (see \cite{clements2008comparison}).
@@
 We argue that this approach should be considered harmful.
-
-\subsection{Library-Oriented Approaches}
-\subsubsection{Embedded DSLs}
-Embedded domain-specific languages are languages that creatively repurpose existing language constructs to create interfaces that resemble those of a distinct language.
+%
+%\subsection{Library-Oriented Approaches}
+%\subsubsection{Embedded DSLs}
+%Embedded domain-specific languages are languages that creatively repurpose existing language constructs to create interfaces that resemble those of a distinct language.
 In languages with rich type systems, such as Haskell, this approach can be quite successful (e.g.
@@
 Ultimately, however, this approach is limited by the host system, and as discussed in Section 2, thus limits experts who want to express particularly novel constructs, relative to the host language.
-
-\subsubsection{Term Rewriting Systems}
-Many languages and tools allow developers to rewrite expressions according to custom rules.
+%
+%\subsubsection{Term Rewriting Systems}
+%Many languages and tools allow developers to rewrite expressions according to custom rules.
 These can broadly be classified as {\it term rewriting systems}.
@@
 Finally, term rewriting systems focus on rewriting terms to support alternative language semantics but do not intrinsically support extensions that cover the full programming system.
-
-\subsubsection{Active Libraries}
-Active libraries, as proposed by Czarnecki et al.
+%
+%\subsubsection{Active Libraries}
+%Active libraries, as proposed by Czarnecki et al.
 \cite{activelibraries}, ``are not passive collections of routines or objects, as are traditional libraries, but take an active role in generating code''.
 They go on to suggest a number of areas in which the library could interact with the programming system, including optimizing code, checking source code for correctness, reporting domain-specific errors and warnings, and ``rendering domain-specific textual and non-textual program representations and for interacting with such representations''.
-
-This paper was largely a proposal.
+%
+%This paper was largely a proposal.
 The concrete implementations of this concept have largely been term rewriting systems described above.

### 2013-09-17 thesis-proposal/conclusion.tex
updated proposal
Updated beginning of proposal

@@
 
-Monolithic programming systems enforce a dichotomy between \emph{built-in constructs}, which enjoy support throughout the programming system but must be designed and implemented by the system designer in advance, and \emph{user-defined constructs}, which can be distributed as user libraries, but must creatively combine and repurpose the small set of available built-in constructs to express all desired run-time, compile-time and edit-time behaviors.
-This dichotomy has forced researchers and domain experts who want to significantly advance the state-of-the-art in programming systems, particularly in compiled, statically-typed systems where this dichotomy is most salient, to design and develop new languages.
-But this then couples their core innovations with a collection of unrelated design decisions, requires more effort both for the experts developing the innovation and its targeted user community, and leads to intrinsic problems for users developing applications consisting of components written in different languages.
 \newpage

### 2013-09-17 thesis-proposal/problem.tex
Updated beginning of proposal

@@
 % !TEX root = omar-thesis-proposal.tex
-\section{Motivation}\label{motivation}\begin{figure}
+\vspace{-25pt}
+\section{Motivation}\label{motivation}
+Designing and implementing a programming language together with its supporting tools (collectively, a \emph{programming system}) that has a sound theoretical foundation, helps users identify and fix errors as early as possible, supports a natural programming style, and that performs well across diverse problem domains and hardware platforms remains a grand challenge in computing.
+Due to the increasing diversity and complexity of modern problem domains and hardware platforms, it has become clear that no small  collection of general-purpose primitives backed by a simple toolchain will be able to fully satisfy these criteria in all cases.
+Instead,  researchers will continue to provide specialized abstractions, notations,  type systems, implementation strategies, optimizations, run-time systems and tool support (collectively, specialized \emph{features}) in order to meet these growing challenges.
+%, researchers and domain experts must continue to design and develop, in a principled manner, specialized notations, type systems, optimizations, run-time systems and tools (collectively, \emph{features}) to support developers as they work in increasingly complex problem domains and on modern hardware systems.
+
+% Rather, researchers and domain experts must continue, in a principled manner, to specify and implement novel abstractions, provide alternative implementations of existing abstractions, supplement languages with useful tools and improve tools with new behaviors (collectively, realize new \emph{features}) that better satisfy these criteria by balancing the concerns relevant to particular application domains.
+
+Ideally, providers would develop and distribute new features as libraries, so that developers could granularly choose those that best satisfy the constraints of their problem domains.
+Unfortunately, this is often infeasible, because libraries are nearly exclusively vehicles for specifying the run-time behaviors of a program, but providing these kinds of features  requires or benefits from some measure of control over the compile-time or edit-time semantics of the programming system.
+That is, from the perspective of a library, the language's semantics are fixed in advance, the compiler and run-time system are ``black box'' implementations of these fixed semantics, and the other tools, like code editors and debuggers, operate according to domain-agnostic protocols also based exclusively on these fixed semantics.
+As a result, providers of new system features must take \emph{language-external approaches}.
+We will argue that such approaches have several fundamental problems, and that taking them has led to an unnecessary gap between research and practice.
+In place of these approaches, we will advocate for \emph{language-internal extensibility mechanisms}, and show how, by organizing new features around types and constraining them appropriately, such mechanisms can be made both safe and highly expressive.
+%But taking a \emph{language-internal approach} to implementing a feature is the most practical.
+If a feature can be realized by creatively using existing language constructs and distributed as a library, clients face fewer barriers to adoption because it is easy to integrate library-based features into existing projects gradually and granularly and they leverage well-understood and well-developed mechanisms.
+% But taking this approach is often \emph{not} possible today %We call designs \emph{monolithic programming systems}.
+
+%To realize a new abstraction or system behavior, such experts can consider either a \emph{language-internal approach}, where they work within an existing language and distribute their solutions as libraries, or a \emph{language-external approach}, where they create a new, distinct programming system (often centered around what has come to be called a new \emph{domain-specific language} \cite{dsl}) or extend an existing system by some mechanism that is not part of the language itself, such as an extension mechanism supported by a {particular} compiler, editor or other tool.
+\subsection{Motivating Example: Regular Expressions}\label{regex}
+To make the issue concrete, let us begin with a simple example.
+\emph{Regular expressions} are a widely-used mechanism for finding patterns in unstructured and semi-structured strings (when formal grammars are not appropriate) \cite{regex}.
+If a programming system included full compile-time and edit-time support for regular expressions, it might provide features like:
+
+\begin{enumerate}
+\item \textbf{built-in syntax for pattern literals} so that malformed patterns result in intelligible \emph{compile-time} parsing errors, motivated by the frequency of run-time errors relating to pattern syntax in Java \cite{regex-type-system} and other languages that do not have this feature.
+\item \textbf{type checking logic} that ensures that key constraints related to regular expressions are not violated, such as that out-of-bounds group indices are not used \cite{regex-type-system} and that only values with correct types are spliced into regular expressions.
+When a type error is found, the error message should again be intelligible.
+\item \textbf{translation logic} that partially or fully compiles regular expressions into the efficient internal representation that will be used by the regular expression matching engine at run-time.
+In most languages, this compilation step occurs at run-time, even if the pattern is fully known at compile-time, thereby introducing latency into programs.
+If the developer is not careful, regular expressions used repeatedly in a program might be needlessly re-compiled on each use.
+%By performing this step ahead-of-time, these dangers can be avoided.
+\item \textbf{editor-integrated tooling} for interactively testing patterns against example strings, quickly referring to documentation, searching databases of common patterns and other domain-specific edit-time facilities.
+\end{enumerate}
+
+%In a conventional \emph{monolithic} programming system, support for each of these features would need to be built into the language and tools. 
+No system today has built-in support for all of the features enumerated above.
+Instead, libraries generally provide support for regular expressions by leveraging  general-purpose constructs.
+Unfortunately, it is impossible to fully encode the syntax and the specialized static and dynamic semantics described above in terms of standard general-purpose notations and abstractions.
+Library providers have thus needed to compromise, typically by representing regular expressions using strings and deferring parsing, typechecking and translation to run-time, which introduces performance overhead and can lead to unanticipated run-time errors (as shown in \cite{regex-type-system}) and security vulnerabilities (due to injection attacks when splicing is used, for example).
+Similarly, edit-time features, as described above, are rarely integrated into editors, and never in a way that facilitates their discovery and use directly when the developer is manipulating regular expressions.
+In most cases, tools must be discovered independently and accessed externally (for example, via a browser), making both their development and use less common and more awkward than necessary \cite{active-code-completion} (see Section \ref{acc}).
+
+\subsection{Language-External Approaches}
+\begin{figure}
 \begin{center}
@@
 Users can only safely and naturally call into languages if the interface uses common constructs and an interoperability layer has been developed (the \emph{interoperability problem}).
-(b) With the language-internal approach, there is one extensible host language and the compile-time and edit-time logic governing novel constructs is expressed within libraries.
+(b) With the active library approach, there is one extensible host language and the compile-time and edit-time logic governing novel constructs is expressed within libraries.
 If the extension mechanism guarantees the safety of arbitrary compositions of extensions, the necessary primitives can simply be imported by library clients as needed, and so interoperability is not a problem.}
@@
 \end{figure}
-Designing and implementing a programming language together with its supporting tools (collectively, a \emph{programming system}) that has a sound theoretical foundation, helps users identify and fix errors as early as possible, supports a natural programming style, and that performs well across diverse problem domains and hardware platforms remains a grand challenge in computing.
-It has become clear that no small, fixed collection of primitive abstractions and tools can fully satisfy these criteria in all situations.
-Rather, researchers and domain experts must continue, in a principled manner, to design and implement novel abstractions, provide alternative implementations of existing abstractions, and supplement languages with new tools, and tools with new behaviors, to better satisfy these criteria within the constraints of different problem domains.
 
-%To realize a new abstraction or system behavior, such experts can consider either a \emph{language-internal approach}, where they work within an existing language and distribute their solutions as libraries, or a \emph{language-external approach}, where they create a new, distinct programming system (often centered around what has come to be called a new \emph{domain-specific language} \cite{dsl}) or extend an existing system by some mechanism that is not part of the language itself, such as an extension mechanism supported by a {particular} compiler, editor or other tool.
-
-When possible, taking a \emph{language-internal approach} to implementing a new abstraction or system behavior (collectively, a new \emph{feature}) is simpler and more practical.
-If a feature can be realized by creatively repurposing existing features and distributed as a library, both providers and clients face fewer barriers to adoption because it is easy to gradually integrate library-based features into existing code and they leverage well-understood and well-developed mechanisms.
-But taking this approach is often \emph{not} possible today because libraries are generally vehicles for specifying the run-time behavior of a program.
-The compile-time and edit-time behaviors of a programming system cannot easily be adapted or extended in a specialized way from within libraries in most programming systems.
-That is, the language's syntax, static semantics and corresponding dynamic semantics are fixed in advance, the compiler is a ``black box'' implementation of these fixed semantics, and the other tools, like code editors and debuggers, operate according to fixed, domain-agnostic protocols based on these fixed semantics.
-
-If any of these components of the system must be modified to realize a new feature, providers must instead take a \emph{language-external approach}, either by developing a new or derivative programming system (often centered around a so-called \emph{domain-specific language} \cite{dsl}) or by extending an existing system by some mechanism that is not part of the language itself, such as an extension mechanism supported by a {particular} compiler or other tool.
-%Extending a particular tool is morally equivalent, for our purposes, to creating a new system, because we aim to minimize the extent to which the edit-time, compile-time and run-time semantics of libraries depend on the implementation details of a particular tools.
+When the compile-time or edit-time semantics of a system must be modified to fully realize a new feature, as in the example above, providers often take a \emph{language-external approach}, either by developing a new or derivative programming system (often centered around a so-called \emph{domain-specific language} \cite{dsl}) or by extending an existing system by some mechanism that is not part of the language itself, such as an extension mechanism supported by a {particular} compiler\footnote{Note that compilers that modify or allow modification of  the semantics of their base language, rather than simply implementing meaning-preserving optimizations, should be considered as pernicious means for creating new languages.
+Many compilers, including \texttt{gcc}, \texttt{GHC} and \texttt{SML/NJ}, are guilty of this sin, meaning that some programs that seem to be written in C, Haskell or Standard ML are actually written in tool-specific derivatives of those languages.
+Language-internal mechanisms do not lead to such fragmentation.} or other tool.
+%This latter method couples the semantics of the feature to the implementation details of a particular tool.
+Because the use of one implementation entails a different semantics for the feature than another, the extended tool acts, \emph{de facto}, as a distinct system for our purposes.
 
 Unfortunately, when providers of new features take language-external approaches, it causes problems for clients.
-Features become coupled to a collection of other unrelated and often underdeveloped features, making adoption more costly when these other choices are not appropriate.
-For example, although recent evidence suggests that developers prefer language-integrated parallel programming abstractions to library-based implementations if all else is equal \cite{langvslib}, library-based implementations remain more widely adopted because they do not require porting applications to a language designed around a few privileged parallel programming abstractions.
-These abstractions may only be useful in performance-critical portions of the application, and then only in cases where no competing abstraction would be more appropriate.
+Features cannot be adopted individually, but instead are only available coupled to a collection of other often-unrelated  features (the incidental features included in a newly-designed language or tool, or the existing features of the particular tool that was externally extended).
+This makes adoption more costly when these other features are either not appropriate or insufficiently developed, or when the features bundled with a different language or tool are simultaneously desirable.
+For example, although evidence suggests that developers prefer language-integrated parallel programming abstractions to library-based implementations if all else is equal \cite{langvslib}, library-based implementations are more widely adopted because parallel programming languages privilege only a few abstractions, even though parallel programming is rarely the only concern relevant to a component or application.
+Regular expression support, for example, may be simultaneously desirable when processing large amounts of textual data in parallel.
+%Similarly, a language and tools designed primarily to support regular expressions might make an interesting research project, but it would not be a suitable tool for writing large applications with more varied needs.
 
@@
 %Clients seem to prioritize the ability to choose different features for different portions of an application. 
-If it were easy to call between languages, then having a variety of specialized languages would be less of a problem.
-In fact, some have advocated for just this development model, calling it the the \emph{language-oriented approach} to software development \cite{languageoriented}.
-A fundamental problem with this approach, however, arises at the interfaces between languages.
-The specialized abstractions particular to one language cannot be safely and naturally expressed in another, in general, so building a program out of components written in many different languages is difficult or impossible whenever these specialized abstractions are exposed at library interface boundaries.
-We refer to this as the \emph{interoperability problem}.
+%If calling between languages were safe and easy, then using a variety of specialized languages and associated tools might be less problematic. In fact, s
+%Recognizing the limitations of relying on monolithic collection of primitives, some researchers have advocated instead for a model where multiple languages used within a single application, calling it the \emph{language-oriented approach} to software development \cite{languageoriented}.
 
-One strategy increasingly being taken by programming language designers to partially address the interoperability problem today is to target an established language, such as the Java Virtual Machine (JVM) bytecode, and support a superset of its constructs.
-Scala \cite{scala} and F\# \cite{fsharp} are examples of general-purpose languages that have taken this approach.
-This only provides full interoperability in one direction (Figure \ref{approaches}a).
-While calling into the common language becomes straightforward, calls in the other direction, or between the languages sharing the common target, are still restricted by the constructs available in the common language.
-If full interoperability is desired, new languages must only include constructs that can already be expressed safely and reasonably naturally in the common language.
-Many innovative features, however, can be difficult to define in terms of existing features in ways that guarantee all necessary invariants are maintained and that do not require large amounts of boilerplate code.
+Even when various concerns can be separated into different components, each written in a suitable language (often called the \emph{language-oriented approach} to software development \cite{language-oriented}), the interfaces between these components remains an issue.
+The specialized primitives particular to one language cannot always be safely and naturally expressed in terms of those available in another, so building a program out of components written in a variety of different languages is difficult or impossible whenever these primitives are exposed at interface boundaries.
+We refer to this fundamental issue as the \emph{interoperability problem}.
+
+One strategy often taken by programming language designers to partially address the interoperability problem is to  target an established language or bytecode, such as the Java Virtual Machine (JVM) bytecode or LLVM IR, and support a superset of its constructs.
+Scala \cite{scala} and F\# \cite{fsharp} are two prominent examples of general-purpose languages that have taken this approach, and many domain-specific language frameworks also rely on this strategy (e.g.
+Delite \cite{delite}).
+This only enables full interoperability in one direction (Figure \ref{approaches}a).
+While calling into the common language becomes straightforward, calls in the other direction, or between the languages sharing the common target, are still restricted by the semantics of the common language.
+If new languages only include constructs that can already be expressed safely and reasonably naturally in the common language, then this approach can work well. 
+But many of the most innovative constructs found in modern languages (often, the features that justify their creation) are difficult to define in terms of existing constructs in ways that guarantee all necessary invariants are statically maintained and that do not require large amounts boilerplate code and run-time overhead.
 For example, the type system of F\# guarantees that null values cannot occur within F\# data structures, but maintaining this important invariant still requires run-time checks because the typing rules of F\# do not apply when F\# code is called from other languages on the Common Language Infrastructure (CLI) like C\#.
-The F\# type system also includes support for checking that units of measure are used correctly \cite{fsharpunits}, but this domain-specific invariant is left completely unchecked at language boundaries.
+The F\# type system also includes support for checking that units of measure are used correctly \cite{fsharpunits}, but this specialized invariant is left completely unchecked at language boundaries.
 In Scala, interfaces built around traits that have default method implementations are difficult to implement from Java or other JVM languages and the workaround can break if the trait is modified \cite{scalatraitinterop}.
@@
 
-For these reasons, we argue that the language-oriented approach is fundamentally flawed and so taking a language-external approach to realizing a new feature, where new features are packaged into mutually incompatible languages and associated tools, should be considered harmful and avoided when possible.
-The goal of the research proposed here is to fundamentally reorganize the core components of the programming system so that such language-external approaches are less frequently necessary, by designing \emph{language-internal extension mechanisms} that give developers control over edit-time and compile-time behaviors that have previously been centrally-controlled by  the  committees that govern languages or tools\footnote{One might compare today's monolithic programming systems to  {centrally-planned} economies, whereas extensible\- languages more closely resemble modern market economies.}\todo{read Arch D.
+\subsection{Active Libraries}
+For these reasons, we argue that taking a language-external approach to realizing a new feature, by packaging it into a new or derivative language or tool, should be considered harmful and avoided whenever possible.
+The goal of the research being proposed here is to fundamentally reorganize the core components of the programming system so that such language-external approaches are less frequently necessary, by designing \emph{language-internal extension mechanisms} that give developers control over edit-time and compile-time behaviors that have previously been defined externally and thus, in the case of popular languages, become available only subject to the approval of slow-moving governing committees\footnote{One might compare today's monolithic programming systems to  {centrally-planned} economies, whereas extensible\- systems more closely resemble modern market economies.}\todo{read Arch D.
 Robison.
@@
 Specifically, we will show how control over \textbf{parsing}, \textbf{typechecking}, \textbf{translation} (the first stage of compilation) and \textbf{code completion} can be delegated to user-defined logic distributed in {libraries}, as illustrated in Figure \ref{approaches}b.
-Such libraries are called \emph{active libraries} because, rather than being passive consumers of features already available in the system, they contain logic invoked directly by tools to provide new domain-specific functionality during development or compilation \cite{active-libraries}.
-Features implemented within active libraries can be imported as needed by clients of libraries that rely on them, unlike features implemented by language-external means.
+Such libraries are called \emph{active libraries} because, rather than being passive consumers of features already available in the system, they contain logic invoked by the system during development or compilation to provide new domain-specific features  \cite{active-libraries}.
+Features implemented within active libraries are imported as needed by the clients of libraries that rely on them, unlike features implemented by language-external means.
 
-Some critical issues having to do with {safety} must be overcome before library-based extension mechanisms can be introduced into a programming system, because if too much control over such core aspects of the system is given to users, the system may become unreliable.
-For example, an extension could weaken important metatheoretic guarantees previously provided by the system.
-Type safety, for instance, may no longer hold if the static and dynamic semantics of the language can be modified arbitrarily within libraries.
-Furthermore\-, even if extensions can be shown not to cause such problems in isolation, there may still be conflicts between extensions that could lead to subtle problems at link-time.
+Some critical issues having to do with {safety} must be overcome before library-based extension mechanisms can be introduced into a programming system, because if too much control over such core aspects of the system is given  to developers, the system may become unreliable.
+%For example, an extension could weaken important metatheoretic guarantees previously provided by the system. 
+Type safety, for example, may not hold if the static and dynamic semantics of the language can be modified or extended arbitrarily from within libraries.
+Furthermore, even if extensions can be shown not to cause such problems in isolation, there may still be conflicts between extensions that could weaken their semantics and lead to subtle problems at link-time.
 For example, if two active libraries defined differing semantics for the same syntactic form, the issue would only manifest itself when both libraries were imported somewhere within the same program.
-These kinds of safety issues have plagued previous attempts at language-internal extensibility (discussed further in the next section).
-To prevent them, the system must strictly constrain extensions such that they are guaranteed safe in isolation and also safely composable in any combination.
+These kinds of safety issues have plagued previous attempts at language-internal extensibility.% To prevent them, our mechanisms will organize  extension logic around types to guarantee that extensions are both safe in isolation and also safely composable in any combination.
 
-The mechanisms described in this thesis are designed to be highly \textbf{expressive}, allowing library-based implementations of features comparable to built-in features found in modern programming systems, without compromising on these important notions of \textbf{safety}.
-This is accomplished by organizing extension logic around types, rather than languages.
-By scoping new compile-time and edit-time logic to expressions of the type with which it is associated, rather than applying it globally or only within contiguous syntactic blocks as in many previously-developed mechanisms, conflicts are avoided.
-We call types with such logic associated with them \emph{active types} and systems that support them \emph{actively-typed programming systems}.
-By constraining the extension logic with a type system and using techniques from the compiler correctness literature, we can ensure that the system as a whole maintains important safety properties.
  %This represents a minimalist approach to system design -- the conventional distinction between built-in and user-defined constructs is blurred and most features of the system are orthogonally implemented as {libraries}, rather than by the maintainers of the system.

### 2013-09-17 thesis-proposal/timeline.tex
Updated beginning of proposal

@@
-\section{Research Plan}
-
-\subsection{Spring 2013}
-I will complete the work on Ace and active type theory and submit it for publication.
-
-\subsection{Summer 2013}
-I will complete the work on active code prediction and submit it for publication.
+% !TEX root = omar-thesis-proposal.tex
+\section{Timeline}
 
 \subsection{Fall 2013}
+I will complete the work on Ace (ESOP - Oct.) and active type theory (PLDI - Nov.) and Wyvern (ECOOP - Dec.) and submit each for publication.
+
+Ace:
+- Finish implementation
+- Finish OpenCL implementation
+- Finish C implementation
+- Implement global arrays, functional data parallelism, objects
+
+Theory:
+- Prove metatheory (mechanized?)
+- Figure out composition theorem
+- Figure out how it would look in Coq
+
+Wyvern:
+- Figure out grammar and type system formalism
+- See implementation through
+- Declarative stuff
+- Write it up
+
+Graphite:
+- Figure out safety stuff
+
+\subsection{Spring 2014}
 I will finalize all work and write the final dissertation.
+
+%\subsection{Summer 2014}

### 2013-10-09 esop14/att-esop14.tex
ESOP14 update

\documentclass{llncs}
\usepackage{llncsdoc}
\usepackage{mathpartir}
\usepackage{amsmath}
\usepackage{amssymb} 
%\usepackage{amsthm}
\usepackage{ stmaryrd }
%\usepackage{cite}
%\renewcommand{\citepunct}{,\,} % IEEEtran wants to use ],\,[ for this but that looks dumb...

\input{macros-atlam}

\renewcommand{\ttdefault}{txtt}
\usepackage{alltt}
\usepackage{listings}
\lstset{language=ML,
showstringspaces=false,
basicstyle=\ttfamily\footnotesize,
morekeywords={newcase,extends}}

\usepackage{float}
\floatstyle{ruled}
\newfloat{codelisting}{tp}{lop}
\floatname{codelisting}{Listing}

\usepackage{url}
% url.sty was written by Donald Arseneau. It provides better support for
% handling and breaking URLs. url.sty is already installed on most LaTeX
% systems. The latest version can be obtained at:
% http://www.ctan.org/tex-archive/macros/latex/contrib/misc/
% Read the url.sty source comments for usage information. Basically,
% \url{my_url_here}.

\usepackage{placeins}

%\lefthyphenmin=4
\hyphenation{op-tical net-works semi-conduc-tor}

\usepackage{todonotes}

\begin{document}
\mainmatter  % start of an individual contribution

% first the title is needed
\title{Active Typechecking and Translation: A Safe Language-Internal Extension Mechanism}

% a short form should be given in case it is too long for the running head
\titlerunning{Lecture Notes in Computer Science: Authors' Instructions}

% the name(s) of the author(s) follow(s) next
%
% NB: Chinese authors should write their first names(s) in front of
% their surnames. This ensures that the names appear correctly in
% the running heads and the author index.
%
\author{Cyrus Omar%
\and Jonathan Aldrich}
%
\authorrunning{Lecture Notes in Computer Science: Authors' Instructions}
% (feature abused for this document to repeat the title also on left hand pages)

% the affiliations are given next; don't give your e-mail address
% unless you accept that it will be published
\institute{Carnegie Mellon University, Pittsburgh, PA 15213, USA\\
\texttt{\{comar,aldrich\}@cs.cmu.edu}
}

%
% NB: a more complex sample for affiliations and the mapping to the
% corresponding authors can be found in the file "llncs.dem"
% (search for the string "\mainmatter" where a contribution starts).
% "llncs.dem" accompanies the document class "llncs.cls".
%

\toctitle{Lecture Notes in Computer Science}
\tocauthor{Authors' Instructions}
\maketitle

\begin{abstract}
Researchers and domain experts often propose new language primitives as extensions to the semantics of an existing language. 
But today's statically-typed languages are {monolithic}: they do not expose language-internal mechanisms for implementing the static and dynamic semantics of new primitive types and their associated operators directly, so these experts must instead create new standalone languages.
This causes problems for potential users because building applications from components written in many different languages can be both unsafe and unnatural. 
An {internally-extensible} language could  address these issues, but designing a mechanism that is expressive while maintaining safety remains a challenge. 
Extensions must be modularly verified, their use in any combination must not weaken the metatheoretic properties of the language, nor can they interfere with one another.
We introduce a mechanism called active type-checking and translation (AT\&T) that aims to directly address these issues while remaining highly expressive. 
AT\&T leverages type-level computation, typed compilation techniques and a form of type abstraction to enable library-based implementations of a variety of primitives over a flexible  grammar in a safe manner. 
%In a language supporting AT\&T, primitive types families are represented uniformly at the type level. Each type is equipped with operators, defined using type-level functions, that are invoked specifically during the type-checking and translation of operations relevant to that type. 
%We discuss three points in the design space: (1) a simple judgmental construction of AT\&T designed to distill the essential concepts and admit formal safety theorems, (2) a "real" language called Ace that we use to demonstrate the expressive power of AT\&T in practice, and (3) a language design called Birdie that lifts and extends CoqÍs rich system of dependent types into the type-level, allowing extension developers to fully prove the correctness of extensions statically.
\keywords{extensible languages; active libraries; typed compilation; type-level computation; type abstraction}
\end{abstract}

\section{Motivation}
When designing and implementing a new abstraction, experts typically begin by attempting to define its new constructs in terms of existing language constructs.
By leveraging the {general-purpose} abstraction mechanisms available in modern languages, such as inductive datatypes and object systems, this approach can be quite effective.  
For example, the Delite framework leverages Scala's powerful general-purpose mechanisms to enable a variety of interesting \emph{embedded domain-specific languages} \cite{delite}. 
Unfortunately, there remain situations of interest where general-purpose abstractions fall short. 
For instance, it is difficult to adequately encode advanced type systems in terms of the simpler rules governing general-purpose abstractions (e.g. reasoning about units of measure requires built-in language support in F\# \cite{units-of-measure}). 
Moreover, an encoding must not only be correct, but also sufficiently concise and natural. Regular expressions encoded using inductive datatypes, for example, are often considered overly verbose, so most functional languages support them via strings, which is less safe. \todo{sentence about custom error messages}
Finally, general-purpose abstractions are implemented in a uniform manner. Domain knowledge is thus not applied to eliminate overhead or perform optimizations, and implementations designed for typical application workloads may not be satisfactory in parts of a program where performance is a key criteria, particularly when targeting heterogeneous hardware platforms (e.g. programmable GPUs) and distributed computing resources.

% are at times impossible or impractical. In our example of adding products or sums to Godel's T, although Church encodings are possible \cite{pfpl}, they require a reasonable level of creativity\footnote{Anecdotally, Church encodings are among the more difficult-to-explain topics covered in our undergraduate programming languages course.}. Moreover, they will not offer the same static safety guarantees as a primitive encoding, they are more verbose and they will incur performance overhead by their use of closures rather than a more direct representation. This is not only a problem for simple languages like Godel's T. Several Haskell-based embedded DSLs have also needed to make significant compromises at times \cite{haskellDSLs}. {\color{red} examples? Scala?}

%creating a new language. If this is not practical, the best one can attempt to do is encode the new types in terms of existing types (by a Church encoding, for example). This is generally unsatisfactory --

%Languages implemented using these common patterns are central planning by a language designer or design committee.

Researchers or domain experts who run into situations like these, where more direct control over a language's semantics and implementation are needed, have little choice but to realize new abstractions by creating a new language of some form. They might develop a new standalone language from scratch, modify an implementation of an existing language, or use tools like compiler generators, DSL frameworks and language workbenches \cite{fowler2010domain}. 
%In our simple scenario, we may simply fork our implementation of Godel's T or even edit it directly (a pernicious technique for implementing a new language where the prior one is overwritten). 
%In a more complex scenario, we may instead employ a tool like a compiler generator or DSL framework \cite{fowler2010domain} that can generate a standalone implementation from declarative specifications of language constructs. Some of these tools allow you to package and reuse these specifications (with the important caveat that not all combinations of constructs are valid and free of conflicts, an important modularity issue that we will return to several times in this paper).
The increasing sophistication and ease-of-use of these tools have led to calls for a {\it language-oriented approach} to software development, where different components of an application are written in different specialized languages \cite{journals/stp/Ward94}. Indeed, a number of software ecosystems are now explicitly designed to support many different languages, both general-purpose and domain-specific, atop a common intermediate language. The Java virtual machine (JVM), the Common Language Infrastructure (CLI) and LLVM are prominent examples of such ecosystems.

Unfortunately, this leads to a critical problem at language boundaries: a library's external interface must only use constructs that can reasonably be expressed in \emph{all possible client languages}. This restricts the design of  languages by precluding constructs that rely on statically-checked invariants stronger than those supported by their underlying implementation in the common intermediate language\todo{F\# example?}. At best, constructs like these can be exposed by generating a wrapper where run-time checks have been inserted to guarantee these invariants. This compromises both verifiability and performance. %requires the development of an interoperability layer for every pair of DSLs. 
Moreover, this approach exposes the internals of an implementation to clients, making the abstraction awkward to work with and causing code breakage when implementation details change. This defeats a primary purpose of high-level programming languages: hiding low-level details from clients of an abstraction. We diagram this fundamental \emph{interoperability problem} in Figure \ref{approaches}(a).
\begin{figure}[t]
\vspace{-15px}
\includegraphics[scale=0.415]{approaches.pdf}
\vspace{-30px}
\caption{(a) With the language-oriented approach, different primitive abstractions are packaged into separate languages that extend and target a common intermediate language (e.g. JVM bytecode). Users can only interface with libraries written in another language via the constructs in the common language, causing \emph{interoperability problems}. (b) With the language-internal approach, the semantics of new abstractions (i.e. the logic that governs typechecking and translation to a fixed {internal language}, here labeled \texttt{I}) can be implemented directly within so-called \emph{active libraries}. Clients can import and use these abstractions directly whenever needed.}
\vspace{-10px}
\label{approaches}
\end{figure}

%As a result, domain-specific languages and new general-purpose abstractions alike have experienced relatively slow adoption in practice.
%
%Porting large codebases to new languages is difficult, and the dominant programming languages innovate slowly, so programming language.
%
%More specifically, such languages are neither \emph{internally extensible} because the language itself exposes only natural numbers and functions to its users, nor are they \emph{externally extensible} because no new behaviors can be added to the language's  implementation in a separate module from the one containing the initial implementation.

%This is the essence of a monolithic language implementation: it is impossible for anyone to modularly extend languages defined in this way.

%Programming languages are typically designed around a monolithic collection of primitive type families and operators. Consider, as a simple example, Godel's T \cite{pfpl}, a typed lambda calculus with recursion on primitive natural numbers\todo{add statics to Appendix A}. Although a language designer may casually speak of ``extending Godel's T with primitive product and sum types'', adding these type families and associated operators to this kind of language from within is impossible. That is, Godel's T is not \emph{internally extensible}.

\emph{Internally-extensible programming languages} promise to avoid these problems by providing researchers and domain experts with a mechanism for implementing the semantics of new primitive constructs directly within libraries\todo{mention active libraries}.
%Developers of libraries need only determine whether they are appropriate for their domain, without also considering whether these constructs can be exposed in terms of abstractions appropriate to client code. 
Because primitive constructs are granularly available, rather than packaged into monolithic collections, clients can simply import any necessary primitive constructs when using code that relies on them, and thus achieve full safety, ease-of-use and performance without requiring wrappers or glue code. Providers of components thus need only consider whether primitives that they use are appropriate for their domain, without also considering whether their code might be used in a context where these primitives are not available. We diagram this competing approach in Figure \ref{approaches}(b).

%Researchers and domain experts thus gain the ability to distribute new ideas for evaluation to a broader development community without requiring the approval of maintainers of mainstream languages, large-scale porting of code or explicit interoperability layers.

For a language-internal extension mechanism to be feasible, however, it must achieve expressiveness while also ensuring that extensions cannot compromise the safety properties of the language and its tools, nor interfere with one another. That is, extensions cannot simply be permitted to add arbitrary logic to the type system or compiler, because this would make it possible to break  type safety, decidability or adequacy theorems that are critical to the operation of the language, the compiler or other extensions. We review some previous attempts at language extensibility, and highlight how they do not adequately achieve both safety and expressiveness, in Section \ref{related-work}.
%{\color{red} transition here} Correctness properties of an extension itself should be modularly verifiable, so that its users can rely on it for verifying and compiling their own code. The mechanism must also ensure that desirable metatheoretic properties and global safety guarantees of the language cannot be weakened by extensions. And with multiple independently-developed extensions used at once, the mechanism must further guarantee that they cannot interfere with one another.

In this paper, we introduce a language-internal extensibility mechanism called \emph{active typechecking and translation} (AT\&T) that allows developers to introduce and implement the logic governing new primitive type families and operators from within libraries. 
We argue that this can be accomplished by enriching the type-level language, rather than introducing a separate metalanguage into the system. 
To make this proposal concrete, we begin by introducing a simple core calculus, called \atlam~(for the ``actively-typed lambda calculus''), in Section \ref{atlam}. 
This calculus uses type-level computation of higher kind, along with techniques borrowed from the typed compilation literature and a form of type abstraction that ensures that the implementation details of an extension are not externally visible to guarantee the safety of the language, the decidability of typechecking and compilation and composability of extensions.

In Section \ref{examples}, we show that despite these constraints, this mechanism is expressive enough to admit, within libraries, a number of general-purpose and domain specific abstractions that normally require built-in language support. Our core calculus uses a uniform abstract syntax for primitive operators to simplify our presentation and analysis, but this syntax would be too verbose to be practical. Thus, we begin this section by showing how a key design choice made in the calculus -- to associate operators with type families, forming what we call \emph{active types} -- enables a novel type-directed desugaring mechanism that permits the use of conventional concrete syntax for language extensions.

Our choice of a simply-typed, simply-kinded calculus where expressions are given meaning by translation to a simply-typed internal language appears to occupy a ``sweet spot'' in the design space, and relates closely to how simply-typed functional languages like  ML and Haskell are specified and implemented today. In Section \ref{design}, we briefly discuss other points in the design space of actively-typed languages and describe the sorts of abstractions that the mechanism as we have introduced it is not capable of expressing, suggesting several directions for future research. We conclude with a discussion of related work in Section \ref{related-work}.
%dependently-typed and object-oriented type-level languages, as well as the constraints governing the design of the internal language.  
%We will also note how object-oriented techniques may also be suit, because of a fundamental connection to the \emph{expression problem} \cite{expression-problem}.

%specify new typechecking rules and translation logic from within libraries. The AT\&T mechanism utilizes type-level computation of higher kind and integrates typed compilation techniques into the language to provide strong safety guarantees, while remaining straightforward and expressive.

%AT\&T is general with respect to many choices about the type-level language, the typed internal language and syntax. Choices along these dimensions can affect both expressiveness and ease-of-use. We will begin in Sec. 2 by introducing a minimal system called $@\lambda$ (the ``actively-typed lambda calculus'') that distills the essence of the mechanism in a simply-typed, simply-kinded setting. This will allow us to fully and concisely formalize the language and compiler and give several key safety theorems. We will then continue in Sec. 3 by discussing variants of this mechanism based on other basic paradigms, considering dependently-typed functional languages and object-oriented languages, discussing trade-offs between expressivity and safety when doing so. We have developed a simple prototype called Ace and have used it to develop a number of full-scale language extensions as libraries. We will briefly discuss this language and these extensions in Sec. 4.

%We note at the outset that AT\&T focuses on extending the static semantics of languages with fixed, though flexible, syntax. Language-internal syntax extension mechanisms have been developed in the past (e.g. SugarJ \cite{sugarj}) but they have also suffered from safety problems because grammar composition is not always safe when done in an  unconstrained manner. Constrained approaches that provide stronger safety guarantees have recently been outlined (e.g. Wyvern \cite{globaldsl13}) but we will leave integration of syntax extensions with semantic extensions as future work.
\section{From Extensible Compilers to Extensible Languages}\label{evolution}
To understand the genesis of our internal extension mechanism, it is helpful to begin by considering why most implementations of programming languages cannot even be  externally extended. 
Let us consider, as a simple example, an implementation of G\"odel's T, a typed lambda calculus with recursion on primitive natural numbers  \cite{pfpl}\todo{put statics in Appendix?}. 
A compiler for this language written using a functional language will invariably represent the primitive type families and operators of the object language using {closed} inductive datatypes. 
For example, a simple implementation in Standard ML may be based around these datatypes:
\begin{lstlisting}
  datatype Type = Nat | Arrow of Type * Type
  datatype Exp = Var of var 
               | Lam of var * Type * Exp
               | Ap of Exp * Exp 
               | Z | S of Exp 
               | Natrec of Exp * Exp * Exp
\end{lstlisting}

The logic governing typechecking and translation to an intermediate language (which suitable for optimization by later phases of compilation) will be implemented by exhaustive case analysis over the constructors of \lstinline{Exp}.

In an object-oriented implementation of Godel's T, we might instead encode types and operators as subclasses of abstract classes \lstinline{Type} and \lstinline{Exp}. Typechecking and translation will proceed by the ubiquitous \emph{visitor pattern} \cite{visitor} by dispatching against a set of {known} subclasses of \lstinline{Exp}.

In either case, we encounter the same basic issue: there is no way to modularly extend the representation of primitive type families and operators and implement their associated typechecking and translation logic. 
This issue is related to the widely-discussed \emph{expression problem} (we do not consider the case of adding new functions entirely here) \cite{wadler-expression}.

A number of language mechanisms have been proposed that allow new cases to be added to datatypes and the functions that operate over them in a modular manner. 
In functional languages, we might use \emph{open datatypes} \cite{open-datatypes}. For example, if we wish to extend G\"odel's T with product types, a language supporting open datatypes might allow you to add types and operators for products to the above definitions, if they have been declared open, like this: 
\begin{lstlisting}
  newcase Prod of Type * Type extends Type
  newcase Pair of Exp * Exp extends Exp    (* Intro *)
  newcase PrL of Exp extends Exp           (* Elim Left *)
  newcase PrR of Exp extends Exp           (* Elim Right *)
\end{lstlisting}

The logic for functionality like typechecking and translation might then be specified for only the new cases. For example, a function \lstinline{typeof} that assigns a type to an expression could be extended to support the \lstinline{PrL} operator like so:
\begin{lstlisting}
  typeof PrL(e) = 
    case typeof e of 
      Prod(t1, _) => t1 
    | _ => raise TypeError("<appropriate error message>")
\end{lstlisting}

If we allow users to define new modules containing definitions like these and link them into our compiler, we have succeeded in creating an externally-extensible language implementation, albeit one where safety is not guaranteed (we will return to this point shortly). We have not, however, created an extensible programming language, because other implementations of the language will not necessarily support the same mechanism. 
If our newly-introduced constructs are exposed at a library's  interface boundary, clients using different compilers face the same problems with interoperability that those using different languages face. That is, \textbf{extending a language by extending a single implementation of it is equivalent to creating a new language}. Several prominent language ecosystems today are in a state where a prominent compiler has introduced or supported the introduction of extensions that many libraries have come to rely on, including the Glasgow Haskell Compiler, SML/NJ and the GNU compilers for C and C++. We argue that this practice should be considered harmful.

A more appropriate and useful place for extensions like this is directly within libraries. To enable this, the language must include an appropriately-constrained mechanism to introduce new type families, like \lstinline{Prod}, and operators, like \lstinline{Pair}, \lstinline{PrL} and \lstinline{PrR}, and implement the associated typechecking and translation logic. When encountering these new operators in expressions, the compiler must effectively  hand control over typechecking and translation to user-defined logic. Because this mechanism is {language-internal}, all compilers must support it to satisfy the language specification, and extensions are packaged directly within  libraries.

Statically-typed languages typically make a distinction between \emph{expressions}, which describe run-time computations, and static constructs like types and datatype declarations. The design described above suggests we may now need to add another layer to our language, an {extension language}, where extensions can be declared and implemented. In fact, we will show that \textbf{the natural place for type system extensions is within the type-level language}. The intuition is that extensions to a language's static semantics will need to manipulate types as values at compile-time. Many languages already allow users to write functions over types, effectively manipulating them as values at compile-time (see Sec. \ref{related-work} for examples). The type-level language is often constrained by its own type system (where the types of type-level values are called \emph{kinds} for clarity) that prevents type-level functions from causing problems during compilation. This is precisely the structure that a distinct extension layer would have, and so we will show that it is quite naturally to unify the two in this work.

\section{\atlam}\label{atlam}
\begin{figure}[t]
\small
$$\begin{array}{rccl}	
\textbf{programs} & \rho & ::= & \pfam{\familyDf}{\progsort} \pipe \pdef{t}{\kappa}{\tau}{\progsort} \pipe e\\
\text{primitive ops}		&	\theta	&	::= &	\topsempty \pipe 
												\topp{\theta}{\tops{op}{\kappai}{i}{a}{\tau}}\\
\\
\textbf{expressions} 				&	e	&	::=	&	\evar{x} \pipe 
														\elam{\evar{x}}{\tau}{e} \pipe 
														\eop{Fam}{op}{
															\taui
														}{
  												    		\splat{e}{1}{n}
														} \\
									& 		&		& 	\\

\hspace{-5pt}\textbf{type-level terms} 	& \tau 	& ::= 	& 	\tvar{t} \pipe 
														\tlam{t}{\kappa}{\tau} \pipe 
														\tapp{\tau_1}{\tau_2} \pipe
														\tnil{\kappa} \pipe \tcons{\tau_1}{\tau_2} \pipe 
									                     \tfold{\tau_1}{\tau_2}{x}{y}{\tau_3}
														\\

\text{type-level data}	 			& 		& \pipe	& 	 \iintlit \pipe \iop{\tau_1}{\tau_2} \pipe \tstr{str} \pipe \tunit \pipe 
														\tpair{\tau_{1}}{\tau_{2}} \pipe 
														\tfst{\tau} \pipe 
														\tsnd{\tau} 
														\\
\text{structural equality}  & & \pipe & 					\tifeq{\tau_{1}}{\tau_{2}}{\kappa}{\tau_{3}}{\tau_{4}} 
														\\

\text{types} 						& 		& \pipe	& 	\ttypestd \pipe \tfamcase{\tau}{Fam}{x}{\tau_1}{\tau_2}\\
						%				& & \pipe & \tfamcase{\tau}{Fam}{x}{\tau_1}{\tau_2}\\

\text{denotations} 				& 		 & 	\pipe	&	\tden{\tau_2}{\tau_1} \pipe \terr \pipe \tdencase{\tau}{y}{x}{\tau_1}{\tau_2}\\
 %& & \pipe & 
%														\tdencase{\tau}{y}{x}{\tau_1}{\tau_2}
%														 \\

\text{quoted IL}		&		&	\pipe	&	\titerm{\gamma} \pipe \titype{\sigma} \\

												\\
\textbf{kinds} 					& \kappa	&	::=	&	\karrow{\kappa_1}{\kappa_2} \pipe \klist{\kappa} \pipe \dint \pipe
											    \kstr \pipe
												\kunit \pipe 
												\kpair{\kappa_{1}}{\kappa_{2}} \pipe 
												\kTypeBlur \pipe \kDen \pipe 
												\kIType \pipe \kITerm
												\\
\\												
%\textbf{ops signature}			& \Theta	&	::=	&	\kOpEmpty \pipe \kOp{\Theta}{op}{\kappai}\\
%											 							&		&		&	\\
\textbf{internal terms} 				& 	\gamma	&	::=	&	\evar{x} \pipe 
												\ilam{\evar{x}}{\sigma}{\gamma} \pipe 
												\iapp{\gamma_{1}}{\gamma_{2}} \pipe
												\ifix{\evar{f}}{\sigma}{\gamma} \pipe
												\ipair{\gamma_{1}}{\gamma_{2}} \pipe 
												\ifst{\gamma} \pipe
												\isnd{\gamma}  
												\\
							& 		& 	\pipe	& 
												\iintlit \pipe \iop{\gamma_{1}}{\gamma_{2}} \pipe \iIfEq{\gamma_{1}}{\gamma_{2}}{\dint}{\gamma_{3}}{\gamma_{4}}  
												\\
 &  & \pipe & \tvalof{\tau_2}{\tau_1} \pipe \iup{\tau} \\
%\text{deabstracted}& \gamma & ::= & \mathcal{G}[\gamma, \sigma]\\
\textbf{internal types}			&	\sigma	&	::=	&    \darrow{\sigma_1}{\sigma_2} \pipe
												\dint \pipe
												\dpair{\sigma_1}{\sigma_2} \pipe
												\trepof{\tau} \pipe \dup{\tau}\\
\end{array}$$
\vspace{-10pt}
\caption{\small Syntax of \atlam. Variables $x$ are used in expressions and internal terms and are distinct from type-level variables, $\tvar{t}$. Names $\fvar{Fam}$ are family names (we assume that unique family names can be generated by some external mechanism) and $\opvar{op}$ are operator names. $\tstr{str}$ denotes string literals, $\iintlit$ denotes integer literals and $\oplus$ stands for binary operations over integers. %The productions related to the internal language are written using generators $\mathcal{G}$ and $\mathcal{S}$ to avoid duplicating the syntax of common terms.
\label{grammar}}
\end{figure}
We will now develop a core calculus, called @$\lambda$ for the ``actively-typed lambda calculus'', and discuss how to address safety concerns that arise when giving users this level of control over the semantics of a language and its implementation. 
\subsection{Overview}
\todo{remove $\Theta$ from family}The grammar of \atlam~is shown in Figure \ref{grammar}. The language is a simply-typed lambda calculus with simply-kinded type-level computation. Kinds, $\kappa$, classify type-level terms, $\tau$. 
Types classify expressions, $e$, and are type-level values of kind $\star$ (following System $F_{\omega}$ \cite{fomega}). The type-level language also includes other kinds of terms, such as type-level functions and other type-level data structures. We include lists (required by our mechanism) as well as type-level integers, strings and products for the sake of our examples (see Sec. \ref{termination} for design constraints in general).
%Some kinds of type-level values (including types themselves, but excluding type-level functions) can be compared for structural equality\todo{describe this later?}.

At the top level, programs, $\rho$, consist of a series of declarations followed by an expression. Declarations can be either bindings of type-level terms to type-level variables using \textsf{def} or a declaration of a new primitive type family using \textsf{family} (Sec. \ref{families}). Expressions can be either variables, lambdas, or applications of operators (Sec. \ref{operators}), and are ultimately given meaning by translation to a typed internal language. This language has been chosen, for simplicity, to be a variant of Plotkin's PCF with primitive integers and products, but in practice would include other constructs consistent with its role as a high-level intermediate language. Terms of this language are denoted $\gamma$ and types are denoted $\sigma$. Both of these productions also include additional forms containing type-level terms, $\tau$; these are used during compilation and will be explained below.

\subsection{Example: Natural Numbers}
\begin{figure}[t]
\small
%\begin{lstlisting}
%family Nat of 1 ~ _.\/(int) {
%	z(i:1, a. case len a =_Z 0 then [[Nat<()> ==> 0]] else err);
%	s(i:1, a. get_last_arg a err (\a1. 
%	  dencase a1 of [[t1 ==> v1]] => 
%	    if t1 =_* Nat<()> then [[Nat<()> ==> ^(v1) + 1]] else err
%	  ow err))
%}
%\end{lstlisting}
\begin{flalign}
&\family{Nat}{\kunit}{\\
&\quad\tops{z}{\kunit}{i}{a}{
	\tapp{\tapp{\tvar{no\_args}}{\tvar{a}}}{
		\tlam{\_}{\kunit}{\tden{\titerm{0}}{\ttype{Nat}{\tunit}}}
	}};\\
&\quad\tops{s}{\kunit}{i}{a}{
	\tapp{
		\tapp{\tvar{last\_arg}}{\tvar{a}}}{\tlam{t1}{\kTypeBlur}{\tlam{v1}{\kITerm}{
			\\
			&\quad\quad\tapp{\tapp{\tapp{\tvar{check\_type}}{\tvar{t1}}}{\ttype{Nat}{\tunit}}}{
				\tden{\titerm{\iup{\tvar{v1}}+1}}{\ttype{Nat}{\tunit}}
			}
		}}
		}
	};\\
&\quad\tops{rec}{\kunit}{i}{a}{
	\tapp{\tapp{\tvar{next\_arg}}{\tvar{a}}}{\tlam{t1}{\kTypeBlur}{\tlam{v1}{\kITerm}{\tlam{a}{\klist{\kDen}}{\\
	&\quad\quad\tapp{\tapp{\tvar{next\_arg}}{\tvar{a}}}{\tlam{t2}{\kTypeBlur}{\tlam{v2}{\kITerm}{\tlam{a}{\klist{\kDen}}{\\
	&\quad\quad \tapp{\tapp{\tvar{last\_arg}}{\tvar{a}}}{\tlam{t3}{\kTypeBlur}{\tlam{v3}{\kITerm}{\\
	&\quad\quad \tapp{\tapp{\tapp{\tvar{check\_type}}{\tvar{t1}}}{\ttype{Nat}{\tunit}}}{(\\
	&\quad\quad \tapp{\tapp{\tapp{\tvar{check\_type}}{\tvar{t3}}}{\ttype{Arrow}{(\ttype{Nat}{\tunit},\ttype{Arrow}{(\tvar{t2},\tvar{t2})})}}}{(\\
		&\quad\quad c
	)}}
	}}}}}}}}}}}}
\\
&}{XXX}{i}{\titype{\dint}}
\end{flalign}
\caption{Example: primitive natural numbers and $n$-ary ($n > 1$) products in \atlam.}
\label{example}
\end{figure}
To make our discussion of each of these components of the calculus concrete, we will begin with an example showing how to introduce primitive natural numbers, implemented as integers internally, as a user-defined extension.

\subsection{Central Compilation Judgement}
\begin{figure}[t]
\small
\begin{mathpar}
\inferrule{
	\overbrace{\progOK{\emptyctx}{\fvarCtx_0}{\rho}}^{\text{\normalsize Kind Checking}}\\
	\overbrace{\pcompiles{\fvalCtx_0}{\rho}{\gamma}}^{\text{\normalsize Active Typechecking and Translation}}
}{
	\ptcc{\rho}{\gamma}
}
\end{mathpar}
\vspace{-10pt}
\caption{\small Central Compilation Judgement of \atlam.}
\label{ccj}
\end{figure}
The \emph{central compilation judgement}, shown in Figure \ref{ccj}, captures the two phases of compilation: kind checking and active typechecking and translation. The latter phase requires normalizing type-level terms to type-level values, so the  kind checking phase preceding it will ensure that this process not ``get stuck'', as we will elaborate upon when discussing safety and decidability in Sec. \ref{safety}.

\subsection{Indexed Type Families and Types}\label{families}
Declaring a new primitive indexed type family by using \textsf{family} can be compared to adding a new constructor to the compiler's \lstinline{Type} datatype, as suggested in Section \ref{evolution}. The kind of type-level data that the family is indexed by is specified by $\kappaidx$. For example, a base type like $\tvar{nat}$ can be thought of as being the only type in a family, $\fvar{Nat}$, trivially indexed by the unit value, of kind $\kunit$, while families like $\fvar{Pair}$ might be indexed by a pair of types, of kind $\kpair{\kTypeBlur}{\kTypeBlur}$. A type (that is, a type-level term of kind $\kTypeBlur$) is created by naming a family that has been previously been introduced and providing an index of the appropriate kind. For example, $\ttype{Prod}{\tpair{\tvar{nat}}{\tvar{nat}}}$ is the type of a pair of natural numbers in a context where the families $\fvar{Prod}$ and $\fvar{Nat}$ have been introduced as described above and the type-level variable $\tvar{nat}$ has been bound to the type $\ttype{Nat}{\tunit}$.

\subsection{Indexed Operator Families and Active Types}\label{operators}
We build in lambdas as the only binding structure in the language to simplify the core calculus. All other operators are associated with a type. For example, nat / tuples / records.

Operator families.

\subsection{Active Typechecking and Translation}

\subsection{Abstract Representations}

\begin{figure}[t]
\small
$\fbox{\inferrule{}{\progOKX{\progsort}}}$
~~~$\tvarCtx ::= \emptyctx \pipe \tvarCtxX{t}{\kappa}$~~~
$\fCtx ::= \Sigma_0	 \pipe \fvarCtxX$
\begin{mathpar}
\inferrule[family decl kinding]{
	\fvar{Fam} \notin \fvarCtx\\
	\kEq{\kappaidx}\\
	\opType{\tvarCtx}{\fvarCtxX}{\theta}{\Theta}\\\\
	\tKind{\tvarCtx, \tvar{i}:{\kappaidx}}{\fvarCtxX}{\tau}{\kIType}\\
	\progOK{\tvarCtx}{\fvarCtxX}{\rho}
}{
	\progOKX{\pfam{\familyDf}{\rho}}
}

\inferrule[type-level bind kinding]{
	\tKindX{\tau}{\kappa}\\
	\progOK{\tvarCtxX{t}{\kappa}}{\fvarCtx}{\rho}
}{
	\progOKX{\pdef{t}{\kappa}{\tau}{\rho}}
}

\inferrule[exp kinding]{
	\exprOK{\tvarCtx}{\emptyctx}{\fvarCtx}{e}
}{
	\progOKX{e}
}
\end{mathpar}
$\fbox{$\tKindX{\theta}{\Theta}$}$
\begin{mathpar}
\inferrule[no ops]{ }{
	\tKindX{\topsempty}{\kOpEmpty}
}

\inferrule[ops]{
	\opType{\tvarCtx}{\fvarCtx}{\theta}{\Theta}\\
	\opvar{op} \notin \theta\\
	\tKind{\tvarCtx, \tOfKind{\tvar{i}}{\kappai}, \tOfKind{\tvar{a}}{\klist{\kDen}}}{\fvarCtx}{\tau}{\kDen}
}{
	\opType{\tvarCtx}{\fvarCtx}{\topp{\theta}{\tops{op}{\kappai}{i}{a}{\tau}}}{
	\kOp{\Theta}{op}{\kappai}}
}
\end{mathpar}
$\fbox{$\exprOKX{e}$}$
~~~$\itvarCtx ::= \emptyctx \pipe \itvarCtx, \evar{x}$
\begin{mathpar}
\inferrule[e-var-kind]{ }{
	\exprOK{\tvarCtx}{\itvarCtx, \evar{x}}{\fvarCtx}{\evar{x}}
}

\inferrule[e-lam-kind]{
	\tKindX{\tau}{\kTypeBlur}\\
	\exprOK{\tvarCtx}{\eivarCtxX{x}}{\fvarCtx}{e}
}{
	\exprOKX{\elam{\evar{x}}{\tau}{e}}
}

\inferrule[e-op-kind]{
	\fvarOfType{Fam}{\kappaidx}{\Theta} \in \fvarCtx\\
	\kOpS{op}{\kappai} \in \Theta\\
	\tKindX{\taui}{\kappai}\\\\
	\exprOKX{e_1}\\
	\cdots\\
	\exprOKX{e_n}
}{
	\exprOKX{\eop{Fam}{op}{\taui}{\splat{e}{1}{n}}}
}
\end{mathpar}
\caption{\small Kinding for programs. Variable contexts $\tvarCtx$ and $\itvarCtx$ obey standard structural properties. Kinding for type-level terms in Figure \ref{kindtl}. \ \label{kindprof}}
\end{figure}
\begin{figure}[t]
\small
$\fbox{\inferrule{}{\pcompilesX{\rho}}}$
~~~$\fvalCtx ::= \fvalCtx_0 \pipe \fvalCtxX{\fvalDf}$
\begin{mathpar}
\inferrule[att-fam]{
	\pcompiles{\fvalCtxX{\fvalDf}}{\rho}{\gamma}
}{
	\pcompilesX{\pfam{\familyDf}{\rho}}
}

\inferrule[att-def]{
	\tEvalX{\tau}{\tau'}\\
	\pcompilesX{\subst{\tau'}{\tvar{t}}{\rho}}
}{
	\pcompilesX{\pdef{t}{\kappa}{\tau}{\rho}}
}

\inferrule[att-exp]{
	\ecompiles{\emptyctx}{\fvalCtx}{e}{\tau}{\ghat}\\
	\eraseX{\ghat}{\gamma}
}{
	\pcompilesX{e}
}
\end{mathpar}
$\fbox{\inferrule{}{\ecompilesX{e}{\tau}{\ghat}}}$
~~~$\etCtx ::= \emptyctx \pipe \etCtxX{x}{\tau}$
\begin{mathpar}
\inferrule[att-var]{ }{
	\ecompiles{\etCtxX{x}{\tau}}{\fvalCtx}{\evar{x}}{\tau}{\evar{x}}
}

\inferrule[att-lam]{
	\tEvalX{\tau_1}{\ttype{Fam}{\tauidx}}\\
	\ecompiles{\etCtxX{x}{\ttype{Fam}{\tauidx}}}{\fvalCtx}{e}{\tau_2}{\ghat}\\\\
	\tauisdel{Arrow}{\fvalCtx}{Fam}{\tauidx}{\dbar}\\
}{
	\ecompiles{\etCtx}{\fvalCtx}{\elam{\evar{x}}{\tau_1}{e}}{\ttype{Arrow}{\tpair{\ttype{Fam}{\tauidx}}{\tau_2}}}{\ilam{\evar{x}}{\dbar}{\ghat}}
}

\inferrule[att-op]{
	\fvalDf \in \fvalCtx\\
	\tops{op}{\kappai}{i}{a}{\tauop} \in \theta\\
	\tEvalX{\taui}{\taui'}\\\\
	\ecompiles{\etCtx}{\fvalCtx}{e_1}{\tau_1}{\ghat_1}~~~~
	\cdots~~~~
	\ecompiles{\etCtx}{\fvalCtx}{e_n}{\tau_n}{\ghat_n}\\
	\text{$\begin{array}{c}
		\left[\text{$\begin{array}{r}
			\taui'/\tvar{i}\\
			\tden{\titerm{\ghat_1}}{\tau_1} :: \cdots :: \tden{\titerm{\ghat_n}}{\tau_n} :: \tnil{\kDen}/\tvar{a}
			\end{array}$}\right]\tauop
		\Downarrow
		\tden{\titerm{\ghat}}{\ttype{Fam'}{\tauidx}}
		\end{array}$}\\
	\tauisdel{Fam}{\fvalCtx}{Fam'}{\tauidx}{\dbar}\\
	\eCtxTogCtx{Fam}{\fvalCtx}{\etCtx}{\gtCtx}\\
	\checkRC{\gtCtx}{Fam}{\fvalCtx}{\ghat}{\dbar}
}{
	\ecompiles{\etCtx}{\fvalCtx}{\eop{Fam}{op}{\taui}{\splat{e}{1}{n}}}{\ttype{Fam'}{\tauidx}}{\ghat}
}
\end{mathpar}
\caption{\small Active typechecking and translation}
\end{figure}
\begin{figure}[t]
\small
\vspace{-15pt}
$\fbox{\inferrule{}{\tKindX{\tau}{\kappa}}}$
\begin{mathpar}
\small
\inferrule[var kind]{
}{
  \tKind{\tvarCtxX{t}{\kappa}}{\fvarCtx}{\tvar{t}}{\kappa}
}

\inferrule[k-arrow intro]{
  \tKind{\tvarCtxX{t}{\kappa_1}}{\fvarCtx}{\tau}{\kappa_2}
}{
  \tKindX{\tlam{t}{\kappa_1}{\tau}}{\karrow{\kappa_1}{\kappa_2}}
}

\inferrule[k-arrow elim]{
  \tKindX{\tau_1}{\karrow{\kappa_1}{\kappa_2}}\\
  \tKindX{\tau_2}{\kappa_1}
}{
  \tKindX{\tapp{\tau_1}{\tau_2}}{\kappa_2}
}
\\
\text{\color{gray} (standard statics for integers, strings, products and lists omitted)}
\\
%%
%%\inferrule{ }{
%%	\tKindX{\tstr{str}}{\kstr}
%%}(\text{str}^I_\tau)
%%
%%\inferrule{ }{
%%	\tKindX{\tunit}{\kunit}
%%}(\text{1}^I_\tau)
%%
%%\inferrule{
%%	\tKindX{\tau_1}{\kappa_1}\\
%%	\tKindX{\tau_2}{\kappa_2}
%%}{
%%	\tKindX{\tpair{\tau_1}{\tau_2}}{\kpair{\kappa_1}{\kappa_2}}
%%}({\times}^I_\tau)
%%
%%\inferrule{
%%	\tKindX{\tau}{\kpair}{\kappa_1}{\kappa_2}
%%}{
%%	\tKindX{\tfst{\tau}}{\kappa_1}
%%}({\times}^{E1}_\tau)
%%
%%\inferrule{
%%	\tKindX{\tau}{\kpair}{\kappa_1}{\kappa_2}
%%}{
%%	\tKindX{\tsnd{\tau}}{\kappa_2}
%%}({\times}^{E2}_\tau)
%%\inferrule{ }{
%%\inferrule{
%%	\tKindX{\tau_{1}}{\kappa_{1}}\\
%%	\tKindX{\tau_{2}}{\kappa_{2}}
%%}{
%%	\tKindX{\tpair{\tau_{1}}{\tau_{2}}}{\kpair{\kappa_{1}}{\kappa_{2}}}
%%}~(\times_\tau)
%%
%%\inferrule{
%%	\tKindX{\tau}{\kpair{\kappa_{1}}{\kappa_{2}}}
%%}{
%%	\tKindX{\tfst{\tau}}{\kappa_{1}}
%%}~(\text{fst}_\tau)
%%
%%\inferrule{
%%	\tKindX{\tau}{\kpair{\kappa_{1}}{\kappa_{2}}}
%%}{
%%	\tKindX{\tsnd{\tau}}{\kappa_{2}}
%%}~(\text{snd}_\tau)
%%
%% TODO: list
\inferrule[eq kind]{
	\kEq{\kappa}\\
	\tKindX{\tau_1}{\kappa}\\
	\tKindX{\tau_2}{\kappa}\\\\
	\tKindX{\tau_3}{\kappa'}\\
	\tKindX{\tau_4}{\kappa'}
}{
	\tKindX{\tifeq{\tau_1}{\tau_2}{\kappa}{\tau_3}{\tau_4}}{\kappa'}
}

\inferrule[type intro]{
	\fvarOfTypeDf \in \fvarCtx\\\\
	\tKind{\tvarCtx}{\fvarCtx}{\tauidx}{\kappaidx}
}{
	\tKindX{\ttype{Fam}{\tauidx}}{\kTypeBlur}
}

\inferrule[type elim]{
	\fvarOfTypeDf \in \fvarCtx\\
	\tKindX{\tau}{\kTypeBlur}\\\\
	\tKind{\tvarCtxX{x}{\kappaidx}}{\fvarCtx}{\tau_0}{\kappa}\\
	\tKindX{\tau_1}{\kappa}
}{
	\tKind{\tvarCtx}{\fvarCtx}{\tfamcase{\tau}{Fam}{x}{\tau_0}{\tau_1}}{\kappa}
}

\inferrule[den intro valid]{
	\tKindX{\tautype}{\kTypeBlur}\\\\
	\tKindX{\tauval}{\kITerm}
}{
	\tKindX{\tden{\tauval}{\tautype}}{\kDen}
}

\inferrule[den intro err]{ }{
	\tKindX{\terr}{\kDen}
}

\inferrule[den elim]{
	\tKindX{\tauden}{\kDen}\\
	\tKind{\tvarCtx, \tOfKind{\tvar{x}}{\kITerm}, \tOfKind{\tvar{y}}{\kTypeBlur}}{\fvarCtx}{\tau_1}{\kappa}\\
	\tKindX{\tau_2}{\kappa}
}{
	\tKindX{\tdencase{\tauden}{y}{x}{\tau_1}{\tau_2}}{\kappa}
}

\inferrule[iterm intro]{
	\isIterm{\tvarCtx}{\emptyctx}{\fvarCtx}{\ghat}
}{
	\tKindX{\titerm{\ghat}}{\kITerm}
}

\inferrule[itype intro]{
	\isItype{\tvarCtx}{\fvarCtx}{\dhat}
}{
	\tKindX{\titype{\dhat}}{\kIType}
}
\end{mathpar}
$\fbox{$\kEq{\kappa}$}$
\begin{mathpar}
\inferrule[t-eq]{ }{
	\kEq{\kTypeBlur}
}

\inferrule[z-eq]{ }{
	\kEq{\dint}
}

\inferrule[str-eq]{ }{
	\kEq{\kstr}
}

\inferrule[u-eq]{ }{
	\kEq{\kunit}
}

\inferrule[prod-eq]{
	\kEq{\kappa_1}\\
	\kEq{\kappa_2}
}{
	\kEq{\kpair{\kappa_1}{\kappa_2}}
}

\inferrule[list-eq]{
	\kEq{\kappa}
}{
	\kEq{\klist{\kappa}}
}
\end{mathpar}
$\fbox{$\isItermX{\ghat}$}$
\begin{mathpar}
\inferrule[i-var kind]{ }{
	\isIterm{\tvarCtx}{\eivarCtxX{x}}{\fvarCtx}{\evar{x}}
}

\inferrule[i-lam kind]{
	\isItypeX{\dhat}\\\\
	\isIterm{\tvarCtx}{\eivarCtxX{x}}{\fvarCtx}{\ghat}
}{
	\isItermX{\ilam{\evar{x}}{\dhat}{\ghat}}
}

\inferrule[i-fix kind]{
	\isItypeX{\dhat}\\\\
	\isIterm{\tvarCtx}{\eivarCtxX{x}}{\fvarCtx}{\ghat}
}{
	\isItermX{\ifix{\evar{f}}{\dhat}{\ghat}}
}
\\
\text{\color{gray} (omitted forms have trivially recursive rules)}\\\vspace{-3pt}
\\
\inferrule[iterm unquote kind]{
	\tKindX{\tau}{\kITerm}
}{
	\isItermX{\iup{\tau}}
}

\inferrule[val from den kind]{
	\tKindX{\tautype}{\kTypeBlur}\\
	\tKindX{\tauval}{\kITerm}
}{
	\isItermX{\tvalof{\tauval}{\tautype}}
}
\end{mathpar}
$\fbox{$\isItypeX{\dhat}$}$
\begin{mathpar}
\inferrule[i-int kind]{ }{
	\isItypeX{\dint}
}

\inferrule[i-prod kind]{
	\isItypeX{\dhat_1}\\\\
	\isItypeX{\dhat_2}
}{
	\isItypeX{\dpair{\dhat_1}{\dhat_2}}
}

\inferrule[i-arrow kind]{
	\isItypeX{\dhat_1}\\\\
	\isItypeX{\dhat_2}
}{
	\isItypeX{\darrow{\dhat_1}{\dhat_2}}
}

\inferrule[itype unquote kind]{
	\tKindX{\tau}{\kIType}
}{
	\isItypeX{\dup{\tau}}
}

\inferrule[rep from type kind]{
	\tKindX{\tau}{\kTypeBlur}
}{
	\isItypeX{\trepof{\tau}}
}
\end{mathpar}
\caption{\small Kinding for type-level terms}
\end{figure}

\begin{figure}
\small
$\fbox{$\tEvalX{\tau}{\tau'}$}$
\begin{mathpar}
\inferrule[tl-lam eval]{ }{
	\tEvalX{\tlam{t}{\kappa}{\tau}}{\tlam{t}{\kappa}{\tau}}
}

\inferrule[tl-ap eval]{
	\tEvalX{\tau_1}{\tlam{t}{\kappa}{\tau}}\\
	\tEvalX{\tau_2}{\tau_2'}\\
	\tEvalX{\subst{\tau_2'}{\tvar{t}}{\tau}}{\tau'}
}{
	\tEvalX{\tapp{\tau_1}{\tau_2}}{\tau'}
}

\text{\color{gray} (standard evaluation rules for integers, strings, products and lists omitted)}

\inferrule[tl-eq eval eq]{
	\tEvalX{\tau_1}{\tau_1'}\\
	\tEvalX{\tau_2}{\tau_1'}\\
	\tEvalX{\tau_3}{\tau_3'}
}{
	\tEvalX{\tifeq{\tau_1}{\tau_2}{\kappa}{\tau_3}{\tau_4}}{\tau_3'}
}

\inferrule[tl-eq eval neq]{
	\tEvalX{\tau_1}{\tau_1'}\\
	\tEvalX{\tau_2}{\tau_2'}\\
	\tau_1' \neq \tau_2'\\
	\tEvalX{\tau_4}{\tau_4'}
}{
	\tEvalX{\tifeq{\tau_1}{\tau_2}{\kappa}{\tau_3}{\tau_4}}{\tau_4'}
}

\inferrule[type eval]{
	\tEvalX{\tauidx}{\tauidx'}
}{
	\tEvalX{\ttype{Fam}{\tauidx}}{\ttype{Fam}{\tauidx'}}
}

\inferrule[famcase eval match]{
	\tEvalX{\tau}{\ttype{Fam}{\tauidx}}\\
	\tEvalX{\subst{\tauidx}{\tvar{x}}{\tau_1}}{\tau_1'}
}{
	\tEvalX{\tfamcase{\tau}{Fam}{x}{\tau_1}{\tau_2}}{\tau_1'}
}

\inferrule[famcase eval fail]{
	\tEvalX{\tau}{\ttype{Fam'}{\tauidx}}\\
	\fvar{Fam} \neq \fvar{Fam'}\\
	\tEvalX{\tau_2}{\tau_2'}
}{
	\tEvalX{\tfamcase{\tau}{Fam}{x}{\tau_1}{\tau_2}}{\tau_2'}
}

\inferrule[den eval]{
	\tEvalX{\tautype}{\tautype'}\\
	\tEvalX{\tauval}{\tauval'}
}{
	\tEvalX{\tden{\tauval}{\tautype}}{\tden{\tauval'}{\tautype'}}
}

\inferrule[err eval]{ }{
	\tEvalX{\terr}{\terr}
}

\inferrule[dencase eval valid]{
	\tEvalX{\tauden}{\tden{\tauval}{\tautype}}\\
	\tEvalX{[\tautype/\tvar{x}, \titerm{\tvalof{\tauval}{\tautype}}/\tvar{y}]\tau_1}{\tau_1'}
}{
	\tEvalX{\tdencase{\tauden}{y}{x}{\tau_1}{\tau_2}}{\tau_1'}
}

\inferrule[dencase eval err]{
	\tEvalX{\tauden}{\terr}\\
	\tEvalX{\tau_2}{\tau_2'}
}{
	\tEvalX{\tdencase{\tauden}{y}{x}{\tau_1}{\tau_2}}{\tau_2'}
}

\inferrule[iterm quote eval]{
	\tiEvalX{\ghat}{\ghat'}
}{
	\tEvalX{\titerm{\ghat}}{\titerm{\ghat'}}
}

\inferrule[itype quote eval]{
	\tiEvalX{\dhat}{\dhat'}
}{
	\tEvalX{\titype{\dhat}}{\titype{\dhat'}}
}
\end{mathpar}
$\fbox{$\tiEvalX{\ghat}{\ghat'}$}$
\begin{mathpar}
\inferrule[i-var eval]{ }{
	\tiEvalX{\evar{x}}{\evar{x}}
}

\inferrule[i-lam eval]{
	\tiEvalX{\dhat}{\dhat'}\\
	\tiEvalX{\ghat}{\ghat'}
}{
	\tiEvalX{\ilam{\evar{x}}{\dhat}{\ghat}}{\ilam{\evar{x}}{\dhat'}{\ghat'}}
}

\inferrule[i-fix eval]{
	\tiEvalX{\dhat}{\dhat'}\\
	\tiEvalX{\ghat}{\ghat'}
}{
	\tiEvalX{\ifix{\evar{f}}{\dhat}{\ghat}}{\ifix{\evar{f}}{\dhat'}{\ghat'}}
}
\\
\text{\color{gray} (omitted forms have trivially recursive rules)}
\\
\inferrule[iterm unquote eval]{
	\tEvalX{\tau}{\tau'}
}{
	\tiEvalX{\iup{\tau}}{\iup{\tau'}}
}

\inferrule[val from den eval]{
	\tEvalX{\tautype}{\tautype'}\\
	\tEvalX{\tauval}{\tauval'}
}{
	\tiEvalX{\tvalof{\tauval}{\tautype}}{\tvalof{\tauval'}{\tautype'}}
}
\end{mathpar}
$\fbox{$\tiEvalX{\sigma}{\sigma'}$}$
\begin{mathpar}
\inferrule[i-int eval]{ }{
	\tiEvalX{\dint}{\dint}
}

\inferrule[i-prod eval]{
	\tiEvalX{\dhat_1}{\dhat_1'}\\
	\tiEvalX{\dhat_2}{\dhat_2'}
}{
	\tiEvalX{\dpair{\dhat_1}{\dhat_2}}{\dpair{\dhat_1'}{\dhat_2'}}
}

\inferrule[i-arrow eval]{
	\tiEvalX{\dhat_1}{\dhat_1'}\\
	\tiEvalX{\dhat_2}{\dhat_2'}
}{
	\tiEvalX{\darrow{\dhat_1}{\dhat_2}}{\darrow{\dhat_1'}{\dhat_2'}}
}

\inferrule[itype unquote eval]{
	\tEvalX{\tau}{\tau'}
}{
	\tiEvalX{\dup{\tau}}{\dup{\tau'}}
}

\inferrule[rep from type eval]{
	\tEvalX{\tau}{\tau'}
}{
	\tiEvalX{\trepof{\tau}}{\trepof{\tau'}}
}
\end{mathpar}
\caption{\small Evaluation semantics for type-level terms}
\end{figure}

\begin{figure}
\small
$\fbox{\inferrule{}{\delfromtau{Fam}{\fvalCtx}{\tau}{\dbar}}}$
~~~$\fvalCtx ::= \fvalCtx_0 \pipe \fvalCtx, \fvalDf$
\begin{mathpar}
\inferrule[abs rep from type]{
	\fval{Fam'}{\theta}{i}{\tau} \in \fvalCtx\\
	\tEvalX{[\tauidx/\tvar{i}]\tau}{\titype{\dhat}}\\
	\ddbarX{\dhat}{\dbar}
}{
	\tauisdel{Fam}{\fvalCtx}{Fam'}{\tauidx}{\dbar}
}
\end{mathpar}
$\fbox{\inferrule{}{\ddbarX{\dhat}{\dbar}}}$
\begin{mathpar}
\inferrule[abs int]{ }{
	\ddbarX{\dint}{\dint}
}

\inferrule[abs arrow]{
	\ddbarX{\dhat_1}{\dbar_1}\\
	\ddbarX{\dhat_2}{\dbar_2}
}{
	\ddbarX{\darrow{\dhat_1}{\dhat_2}}{\darrow{\dbar_1}{\dbar_2}}
}

\inferrule[abs prod]{
	\ddbarX{\dhat_1}{\dbar_1}\\
	\ddbarX{\dhat_2}{\dbar_2}
}{
	\ddbarX{\dpair{\dhat_1}{\dhat_2}}{\dpair{\dbar_1}{\dbar_2}}
}

\inferrule[abs+cancel unquote]{
	\ddbarX{\dhat}{\dbar}
}{
	\ddbarX{\dup{\titype{\dhat}}}{\dbar}
}

\inferrule[abs rep from type visible]{
	\fvalDf \in \fvalCtx\\
	\tEvalX{[\tauidx/\tvar{i}]\tau}{\titype{\dhat}}\\
	\ddbarX{\dhat}{\dbar}
}{
	\ddbarX{\trepof{\ttype{Fam}{\tauidx}}}{\dbar}
}

\inferrule[abs rep from type hidden]{
	\fvar{Fam} \neq \fvar{Fam'}
}{
	\ddbarX{\trepof{\ttype{Fam'}{\tauidx}}}{\trepof{\ttype{Fam'}{\tauidx}}}
}
\end{mathpar}
$\fbox{\inferrule{}{\eCtxTogCtxX{\etCtx}{\gtCtx}}}$
~~~$\gtCtx ::= \emptyctx \pipe \gtCtxX{x}{\dbar}$
\begin{mathpar}
\inferrule[abs empty]{ }{
	\eCtxTogCtxX{\emptyctx}{\emptyctx}
}

\inferrule[abs ctx]{
	\eCtxTogCtxX{\etCtx}{\gtCtx}\\
	\delfromtau{Fam}{\fvalCtx}{\tau}{\dbar}
}{
	\eCtxTogCtxX{\etCtxX{x}{\tau}}{\gtCtxX{x}{\dbar}}
}
\end{mathpar}
$\fbox{\inferrule{}{\checkRCX{\ghat}{\dbar}}}$
\begin{mathpar}
\inferrule[abs i-var]{ }{
	\checkRC{\gtCtxX{x}{\dbar}}{Fam}{\fvalCtx}{\evar{x}}{\dbar}
}

\inferrule[abs i-lam]{
	\ddbarX{\dhat_1}{\dbar_1}\\
	\checkRC{\gtCtxX{x}{\dbar_1}}{Fam}{\fvalCtx}{\gamma}{\dbar_2}
}{
	\checkRCX{\ilam{x}{\dhat_1}{\gamma}}{\darrow{\dbar_1}{\dbar_2}}
}

\inferrule[abs i-ap]{
	\checkRCX{\ghat_1}{\darrow{\dbar_1}{\dbar_2}}\\\\
	\checkRCX{\ghat_2}{\dbar_1}
}{
	\checkRCX{\iapp{\ghat_1}{\ghat_2}}{\dbar_2}
}

\inferrule[abs i-fix]{
	\ddbarX{\dhat}{\dbar}\\
	\checkRC{\gtCtxX{x}{\dbar}}{Fam}{\fvalCtx}{\ghat}{\dbar}
}{
	\checkRCX{\ifix{x}{\dhat}{\ghat}}{\dbar}
}
\\
\text{\color{gray} (standard statics for integers and products omitted)}
\\
%\inferrule[abs int]{ }{
%	\checkRCX{\iintlit}{\dint}
%}
%
%\inferrule[abs op]{
%	\checkRCX{\gamma_1}{\dint}\\
%	\checkRCX{\gamma_2}{\dint}
%}{
%	\checkRCX{\iop{\gamma_1}{\gamma_2}}{\dint}
%}
%
%\inferrule[abs pair]{
%	\checkRCX{\gamma_1}{\dbar_1}\\
%	\checkRCX{\gamma_2}{\dbar_2}
%}{
%	\checkRCX{\ipair{\gamma_1}{\gamma_2}}{\dpair{\dbar_1}{\dbar_2}}
%}
%
%\inferrule[abs fst]{
%	\checkRCX{\gamma}{\dpair{\dbar_1}{\dbar_2}}
%}{
%	\checkRCX{\ifst{\gamma}}{\dbar_1}
%}
%
%\inferrule[abs snd]{
%	\checkRCX{\gamma}{\dpair{\dbar_1}{\dbar_2}}
%}{
%	\checkRCX{\isnd{\gamma}}{\dbar_2}
%}
%
\inferrule[abs if eq]{
	\checkRCX{\ghat_1}{\dint}\\
	\checkRCX{\ghat_2}{\dint}\\\\
	\checkRCX{\ghat_3}{\dbar}\\
	\checkRCX{\ghat_4}{\dbar}
}{
	\checkRCX{\iIfEq{\ghat_1}{\ghat_2}{\dint}{\ghat_3}{\ghat_4}}{\dbar}
}

\inferrule[abs iterm unquote]{
	\checkRCX{\ghat}{\dbar}
}{
	\checkRCX{\iup{\titerm{\ghat}}}{\dbar}
}

\inferrule[abs val from den visible]{
	\tauisdel{Fam}{\fvalCtx}{Fam}{\tauidx}{\dbar}\\
	\checkRCX{\ghat}{\dbar}
}{
	\checkRCX{\tvalof{\titerm{\ghat}}{\ttype{Fam}{\tauidx}}}{\dbar}
}

\inferrule[abs val from den hidden]{
	\fvar{Fam} \neq \fvar{Fam'}\\
	\tauisdel{Fam}{\fvalCtx}{Fam'}{\tauidx}{\dbar}\\
	\checkRCX{\ghat}{\dbar}
}{
	\checkRCX{\tvalof{\titerm{\ghat}}{\ttype{Fam'}{\tauidx}}}{\trepof{\ttype{Fam'}{\tauidx}}}
}
\end{mathpar}
\caption{\small Abstracted internal typing}
\end{figure}
\begin{figure}[t]
\small
$\fbox{\inferrule{}{\eraseX{\dbar}{\sigma}}}$
\begin{mathpar}
\inferrule[deabs int]{ }{
	\eraseX{\dint}{\dint}
}

\inferrule[deabs arrow]{
	\eraseX{\dbar_1}{\sigma_1}\\\\
	\eraseX{\dbar_2}{\sigma_2}
}{
	\eraseX{\darrow{\dbar_1}{\dbar_2}}{\darrow{\sigma_1}{\sigma_2}}
}

\inferrule[deabs prod]{
	\eraseX{\dbar_1}{\sigma_1}\\\\
	\eraseX{\dbar_2}{\sigma_2}
}{
	\eraseX{\dpair{\dbar_1}{\dbar_2}}{\dpair{\sigma_1}{\sigma_2}}
}

\inferrule[deabs  rep from type hidden]{
	\fvalDf \in \fvalCtx\\
	\tEvalX{[\tauidx/\tvar{i}]\tau}{\titype{\dhat}}\\\\
	\ddbarX{\dhat}{\dbar}\\
	\eraseX{\dbar}{\sigma}
}{
	\eraseX{\trepof{\ttype{Fam}{\tauidx}}}{\sigma}
}
\end{mathpar}
$\fbox{\inferrule{}{\eraseX{\ghat}{\gamma}}}$
\begin{mathpar}
\inferrule[deabs i-var]{ }{
	\eraseX{\evar{x}}{\evar{x}}
}

\inferrule[deabs i-lam]{
	\ddbar{\_}{\fvalCtx}{\dhat}{\dbar}\\
	\eraseX{\dbar}{\sigma}\\
	\eraseX{\ghat}{\gamma}
}{
	\eraseX{\ilam{x}{\dhat}{\ghat}}{\ilam{x}{\sigma}{\gamma}}
}

%\inferrule[deabs ap]{
%	\eraseX{\ghat_1}{\ghat_1}\\
%	\eraseX{\ghat_2}{\ghat_2}
%}{
%	\eraseX{\iapp{\ghat_1}{\ghat_2}}{\iapp{\ghat_1}{\ghat_2}}
%}
%
\inferrule[deabs i-fix]{
	\ddbar{\_}{\fvalCtx}{\dhat}{\dbar}\\
	\eraseX{\dbar}{\sigma}\\
	\eraseX{\ghat}{\gamma}
}{
	\eraseX{\ifix{x}{\dhat}{\ghat}}{\ifix{x}{\sigma}{\gamma}}
}
\\
\text{\color{gray} (omitted forms have trivially recursive rules)}
\\
\inferrule[deabs unquote]{
	\eraseX{\ghat}{\gamma}
}{
	\eraseX{\iup{\titerm{\ghat}}}{\gamma}
}

\inferrule[deabs val from den hidden]{
	\eraseX{\ghat}{\gamma}
}{
	\eraseX{\tvalof{\titerm{\ghat}}{\ttype{Fam}{\tauidx}}}{\gamma}
}
\end{mathpar}
\caption{\small Deabstraction rules}
\end{figure}
%
%family RECORD of (string, type) dict {
%  new(idx, args. foldl (fn (x, y) => 
%  get[field : string](idx. case find(idx, field) of SOME t => t | _ => err)

\section{Safety of @$\lambda$}\label{safety}

% that's all folks

\section{Examples}\label{examples}

\section{Design Considerations}\label{design}

\section{Related Work}\label{related-work}
\subsection{Type-Level Computation} %Haskell, Ur and $\Omega$mega
System XX with simple case analysis provides the basis of type-level computation in Haskell (where type-level functions are called type families \cite{Chakravarty:2005:ATC}). Ur uses type-level records and names to support typesafe metaprogramming, with applications to web programming \cite{conf/pldi/Chlipala10}. $\Omega$mega adds algebraic data types at the type-level, using these to increase the expressive power of algebraic data types at the expression level \cite{conf/cefp/SheardL07}. Dependently-typed languages blur the traditional phase separation between types and expressions, so type-level computation is often implicitly used (though not always in its most general form, e.g. Deputy \cite{conf/icfp/ChenX05}, ATS \cite{conf/esop/ConditHAGN07}.)

\subsection{Run-Time Indirection}
{\it Operator overloading} \cite{vanWijngaarden:Mailloux:Peck:Koster:Sintzoff:Lindsey:Meertens:Fisker:acta:1975} and {\it metaobject dispatch} \cite{Kiczales91} are run-time protocols that translate operator invocations into function calls. The function is typically selected according to the type or value of one or more operands. These protocols share the notion of {\it inversion of control} with type-level specification. However, type-level specification is a {\it compile-time} protocol focused on enabling specialized verification and implementation strategies, rather than simply enabling run-time indirection.

\subsection{Term Rewriting Systems}
Many languages and tools allow developers to rewrite expressions according to custom rules. These can broadly be classified as {\it term rewriting systems}. Macro systems, such as those characteristic of the LISP family of languages \cite{mccarthy1978history}, are the most prominent example. Some compile-time metaprogramming systems also allow users to manipulate syntax trees (e.g. MetaML \cite{Sheard:1999:UMS}), and external rewrite systems also exist for many languages.
These facilities differ from type-level specification in one or more of the following ways:

\begin{enumerate}
\item In type-level specification, the type of a value is determined separately from its representation; in fact, the same representation may be generated by multiple types. 
\item We draw a distinction between the metalanguage, used to specify types and compile-time logic, the source grammar, used to describe run-time behavior, and the internal language, used to implement this behavior. Term rewriting systems generally do not draw this distinction. By doing so, each component language can be structured and constrained as appropriate for its distinct role, as we show.
%\item With type-level specification, dispatch to a type-level function occurs implicitly on the basis of the structure of an expression. In contrast, most term-rewriting systems operate by  explicit invocation of a macro or specialized syntax. Some LISP macro systems have explored pattern-based dispatch (e.g. A*\cite{fowler2010domain}, EPP\cite{fowler2010domain}) and macro systems for object-oriented languages, like OpenC++ \cite{fowler2010domain} and OpenJava \cite{fowler2010domain}, do offer a somewhat limited form of operation-based dispatch.
\item Many common macro systems and metaprogramming facilities operate at run-time. Compilers for some forms of LISP employ aggressive compile-time specialization techniques to attempt to minimize this overhead. Static and staged term-rewriting systems also exist (e.g. OpenJava\cite{TatM:OpenJCBMSJ}, Template Haskell\cite{SheardPeytonJones:Haskell-02}, MetaML \cite{Sheard:1999:UMS} and others). 
\end{enumerate}

\subsection{Language Frameworks}
When the mechanisms available in an existing language prove insufficient, researchers and domain experts must design a new language. A number of tools have been developed to assist with this task, including compiler generators, language workbenches and domain-specific language frameworks (cf \cite{fowler2010domain}).

A major barrier to adoption is the fact that interoperability is intrinsically problematic. Even languages which target a common platform, such as the Java Virtual Machine, can only interact using its limited set of primitives. Specialized typing rules are not checked at language boundaries, performance often suffers, and the syntax can be unnatural, particularly for languages which differ significantly from the platform's native language (e.g. Java).

Instead of focusing on defining standalone languages, type-level specification gives greater responsibility in a granular manner to libraries. In this way, a range of constructs can coexist within the same program and, assuming that it can be shown by some method that various constructs are safely composable, be mixed and matched. The main limitation is that the protocol requires defining a fixed source grammar, whereas a specialized language has considerable flexibility in that regard. Nevertheless, as Ace shows, a simple grammar can be used quite flexibly.
\subsection{Extensible Compilers}
An alternative methodology is to implement language features granularly as compiler extensions. As discussed in Section 1, existing designs suffer from the same problems related to composability, modularity\-, safety and security as extensible languages, while also adding the issue of language fragmentation.

Type-level specification can in fact be implemented within a compiler, rather than provided as a core language feature. This would resolve some of the issues, as described in this paper. However, by leveraging type-level computation to integrate the protocol directly into the language, we benefit from common module systems and other shared infrastructure. We also avoid the fragmentation issue.
\subsection{Specification Languages}
Several {\it specification languages} (or {\it logical frameworks}) based on these theoretical formulations exist, including the OBJ family of languages (e.g. CafeOBJ \cite{Diaconescu-Futatsugi01}). They provide support for verifying a program against a language specification, and can automatically execute these programs as well in some cases. The  language itself specifies which verification and execution strategies are used.

Type-determined compilation takes a more concrete approach to the problem, focusing on combining {\it implementations} of different\- logics, rather than simply their specifications. In other words, it focuses on combining {\it type checkers} and {\it implementation strategies} rather than more abstract representations of a language's type system and dynamic semantics. In Section 4, we outlined a preliminary approach based on proof assistant available for the type-level language to unify these approaches, and we hope to continue this line of research in future work.

CANT GUARANTEE THAT SPECIFICATIONS ARE ACTUALLY DECIDABLE 
\section{Discussion}\label{discussion}
\bibliographystyle{abbrv}
\bibliography{../research}
\end{document}

### 2013-10-10 esop14/att-esop14.tex
Updated ESOP14

@@
 &\quad\tops{z}{\kunit}{i}{a}{
-	\tapp{\tapp{\tvar{no\_args}}{\tvar{a}}}{
+	\tapp{\tapp{\tvar{const}}{\tvar{a}}}{
 		\tlam{\_}{\kunit}{\tden{\titerm{0}}{\ttype{Nat}{\tunit}}}
@@
 	\tapp{
-		\tapp{\tvar{last\_arg}}{\tvar{a}}}{\tlam{t1}{\kTypeBlur}{\tlam{v1}{\kITerm}{
+		\tapp{\tvar{pop\_final}}{\tvar{a}}}{\tlam{t1}{\kTypeBlur}{\tlam{v1}{\kITerm}{
 			\\
@@
 &\quad\tops{rec}{\kunit}{i}{a}{
-	\tapp{\tapp{\tvar{next\_arg}}{\tvar{a}}}{\tlam{t1}{\kTypeBlur}{\tlam{v1}{\kITerm}{\tlam{a}{\klist{\kDen}}{\\
-	&\quad\quad\tapp{\tapp{\tvar{next\_arg}}{\tvar{a}}}{\tlam{t2}{\kTypeBlur}{\tlam{v2}{\kITerm}{\tlam{a}{\klist{\kDen}}{\\
-	&\quad\quad \tapp{\tapp{\tvar{last\_arg}}{\tvar{a}}}{\tlam{t3}{\kTypeBlur}{\tlam{v3}{\kITerm}{\\
+	\tapp{\tapp{\tvar{pop}}{\tvar{a}}}{\tlam{t1}{\kTypeBlur}{\tlam{v1}{\kITerm}{\tlam{a}{\klist{\kDen}}{\\
+	&\quad\quad\tapp{\tapp{\tvar{pop}}{\tvar{a}}}{\tlam{t2}{\kTypeBlur}{\tlam{v2}{\kITerm}{\tlam{a}{\klist{\kDen}}{\\
+	&\quad\quad \tapp{\tapp{\tvar{pop\_final}}{\tvar{a}}}{\tlam{t3}{\kTypeBlur}{\tlam{v3}{\kITerm}{\\
 	&\quad\quad \tapp{\tapp{\tapp{\tvar{check\_type}}{\tvar{t1}}}{\ttype{Nat}{\tunit}}}{(\\
-	&\quad\quad \tapp{\tapp{\tapp{\tvar{check\_type}}{\tvar{t3}}}{\ttype{Arrow}{(\ttype{Nat}{\tunit},\ttype{Arrow}{(\tvar{t2},\tvar{t2})})}}}{(\\
-		&\quad\quad c
+	&\quad\quad \tapp{\tapp{\tapp{\tvar{check\_type}}{\tvar{t3}}}{\ttype{Arrow}{(\ttype{Nat}{\tunit},\ttype{Arrow}{(\tvar{t2},\tvar{t2})})}}}{\\
+		&\quad\quad \tden{\iapp{(\ifix{f}{\darrow{\dint}{\trepof{\tvar{t2}}}}{\ilam{x}{\dint}{\\
+		&\quad\quad\quad \iIfEq{x}{0}{\dint}{\iup{\tvar{v2}}}{\iapp{\iapp{\iup{\tvar{v3}}}{(x-1)}}{(\iapp{f}{(x-1)})}}}})}{\iup{\tvar{v1}}}}{\tvar{t2}}
 	)}}
@@
 \\
-&}{XXX}{i}{\titype{\dint}}
+&}{XXX}{i}{\titype{\dint}};\\
+&\pdef{nat}{\kTypeBlur}{\ttype{Nat}{\tunit}}{\\
+&(\elam{plus}{\ttype{Arrow}{(\tvar{nat}, \ttype{Arrow}{(\tvar{nat}, \tvar{nat}))}}}{
+	\elam{two}{\tvar{nat}}{\\
+	&\quad\quad \eopapp{\eopapp{plus}{two}}{two}}})
+}\\
+&\quad \elam{x}{\tvar{nat}}{\elam{y}{\tvar{nat}}{
+	\eop{Nat}{rec}{\tunit}{x; y; \elam{p}{\tvar{nat}}{\elam{r}{\tvar{nat}}{
+	\eop{Nat}{s}{\tunit}{r}
+	}}}}}\\
+&\quad  \eop{Nat}{s}{\tunit}{\eop{Nat}{s}{\tunit}{\eop{Nat}{z}{\tunit}{ }}}
+%{\\&\elet{two}{\tvar{nat}}{\eop{Nat}{s}{\tunit}{\eop{Nat}{s}{\tunit}{\eop{Nat}{z}{\tunit}{ }}}}{\eapp{\eapp{plus}{two}}{two}}}
+%}
+%\elam{x}{\tvar{nat}}{\elam{y}{\tvar{nat}}{\\
+	%&\quad \eop{Nat}{rec}{\tunit}{x; y; \elam{p}{\tvar{nat}}{\elam{r}{\tvar{nat}}{
+	%\eop{Nat}{s}{\tunit}{r}
+	%}}}}
 \end{flalign}
-\caption{Example: primitive natural numbers and $n$-ary ($n > 1$) products in \atlam.}
+\caption{Example: primitive natural numbers in \atlam. Type-level helper functions $\tvar{const}$, $\tvar{pop}$, $\tvar{pop\_final}$ and $\tvar{check\_type}$ are given in Appendix \ref{helper}.}
 \label{example}
@@
 \small
-\vspace{-15pt}
+%\vspace{-15pt}
 $\fbox{\inferrule{}{\tKindX{\tau}{\kappa}}}$
@@
 \inferrule[deabs i-lam]{
-	\ddbar{\_}{\fvalCtx}{\dhat}{\dbar}\\
-	\eraseX{\dbar}{\sigma}\\
+	\ddbar{\_}{\fvalCtx}{\dhat}{\dbar}\\\\
+	\eraseX{\dbar}{\sigma}\\\\
 	\eraseX{\ghat}{\gamma}
@@
 \inferrule[deabs i-fix]{
-	\ddbar{\_}{\fvalCtx}{\dhat}{\dbar}\\
-	\eraseX{\dbar}{\sigma}\\
+	\ddbar{\_}{\fvalCtx}{\dhat}{\dbar}\\\\
+	\eraseX{\dbar}{\sigma}\\\\
 	\eraseX{\ghat}{\gamma}
@@
 \bibliography{../research}
+\appendix
+\section{Appendix}
+\subsection{Helper Functions}\label{helper}
+
 \end{document}

### 2013-10-11 esop14/att-esop14.tex
Up through ATT

@@
 \documentclass{llncs}
+\usepackage{amsmath}
 \usepackage{llncsdoc}
-\usepackage{mathpartir}
-\usepackage{amsmath}
 \usepackage{amssymb} 
@@
 \usepackage{ stmaryrd }
+\usepackage{mathpartir}
 %\usepackage{cite}
@@
 \section{Motivation}
-When designing and implementing a new abstraction, experts typically begin by attempting to define its new constructs in terms of existing language constructs.
-By leveraging the {general-purpose} abstraction mechanisms available in modern languages, such as inductive datatypes and object systems, this approach can be quite effective.  
+When designing and implementing a new abstraction, experts typically begin by attempting to define new constructs in terms of existing language constructs.
+This approach is often effective because modern {general-purpose} abstraction mechanisms, like inductive datatypes and object systems, are highly expressive.  
 For example, the Delite framework leverages Scala's powerful general-purpose mechanisms to enable a variety of interesting \emph{embedded domain-specific languages} \cite{delite}. 
-Unfortunately, there remain situations of interest where general-purpose abstractions fall short. 
+Unfortunately, there remain some situations of interest where general-purpose abstractions fall short. 
 For instance, it is difficult to adequately encode advanced type systems in terms of the simpler rules governing general-purpose abstractions (e.g.
 reasoning about units of measure requires built-in language support in F\# \cite{units-of-measure}).
-Moreover, an encoding must not only be correct, but also sufficiently concise and natural.
-Regular expressions encoded using inductive datatypes, for example, are often considered overly verbose, so most functional languages support them via strings, which is less safe.
-\todo{sentence about custom error messages}
-Finally, general-purpose abstractions are implemented in a uniform manner.
+Even if a full encoding is possible, it may not be useful if it is overly verbose or unnatural, or if the error messages are overly abstract.
+For example, regular expressions encoded using inductive datatypes are often considered overly verbose, so most functional languages support them via strings, which is less safe.
+Finally, general-purpose abstractions are also implemented in a uniform manner.
 Domain knowledge is thus not applied to eliminate overhead or perform optimizations, and implementations designed for typical application workloads may not be satisfactory in parts of a program where performance is a key criteria, particularly when targeting heterogeneous hardware platforms (e.g.
@@
 
-Researchers or domain experts who run into situations like these, where more direct control over a language's semantics and implementation are needed, have little choice but to realize new abstractions by creating a new language of some form.
+Researchers or domain experts who run into situations like these, where more direct control over a language's semantics and implementation are needed, have little choice today but to realize new abstractions by creating a new language of some form.
 They might develop a new standalone language from scratch, modify an implementation of an existing language, or use tools like compiler generators, DSL frameworks and language workbenches \cite{fowler2010domain}.
@@
 Unfortunately, this leads to a critical problem at language boundaries: a library's external interface must only use constructs that can reasonably be expressed in \emph{all possible client languages}.
-This restricts the design of  languages by precluding constructs that rely on statically-checked invariants stronger than those supported by their underlying implementation in the common intermediate language\todo{F\# example?}.
+This discourages languages from including constructs that rely on statically-checked invariants stronger than those supported by their underlying implementation in the common intermediate language.
 At best, constructs like these can be exposed by generating a wrapper where run-time checks have been inserted to guarantee these invariants.
 This compromises both verifiability and performance.
-%requires the development of an interoperability layer for every pair of DSLs.
+For%requires the development of an interoperability layer for every pair of DSLs.
 Moreover, this approach exposes the internals of an implementation to clients, making the abstraction awkward to work with and causing code breakage when implementation details change.
@@
 We diagram this fundamental \emph{interoperability problem} in Figure \ref{approaches}(a).
+As an example, F\#'s type system prevents \lstinline{null} values from occurring within data structures, but because it's type system is not available when calling into F\# code from another language, like C\#, run-time null checks must still be inserted.
 \begin{figure}[t]
@@
 %Developers of libraries need only determine whether they are appropriate for their domain, without also considering whether these constructs can be exposed in terms of abstractions appropriate to client code.
-Because primitive constructs are granularly available, rather than packaged into monolithic collections, clients can simply import any necessary primitive constructs when using code that relies on them, and thus achieve full safety, ease-of-use and performance without requiring wrappers or glue code.
-Providers of components thus need only consider whether primitives that they use are appropriate for their domain, without also considering whether their code might be used in a context where these primitives are not available.
+As a result, clients can granularly import any necessary primitive constructs when using code that relies on them, and thus achieve full safety, ease-of-use and performance due to the absence of wrappers and glue code.
+Providers of components thus need only consider whether primitives that they use are appropriate for their domain, without also considering whether their code might be used in a context where these primitives are not otherwise appropriate.
+Libraries containing logic that is invoked at compile-time have been called \emph{active libraries} \cite{activelibraries}.
 We diagram this competing approach in Figure \ref{approaches}(b).
@@
 
-In Section \ref{examples}, we show that despite these constraints, this mechanism is expressive enough to admit, within libraries, a number of general-purpose and domain specific abstractions that normally require built-in language support.
-Our core calculus uses a uniform abstract syntax for primitive operators to simplify our presentation and analysis, but this syntax would be too verbose to be practical.
-Thus, we begin this section by showing how a key design choice made in the calculus -- to associate operators with type families, forming what we call \emph{active types} -- enables a novel type-directed desugaring mechanism that permits the use of conventional concrete syntax for language extensions.
+In Section \ref{examples}, we show that despite these constraints, this mechanism is expressive enough to admit, within libraries, a number of general-purpose and domain-specific abstractions that normally require built-in language support.
+Our core calculus uses a uniform abstract syntax for primitive operators to simplify our presentation and analysis, but this syntax is too verbose to be practical.
+Thus, we begin this section by showing how a key design choice made in the calculus -- to associate operators with type families, forming what we call \emph{active type families} -- supports a novel type-directed desugaring mechanism that permits the use of conventional concrete syntax for language extensions.
 
@@
 To understand the genesis of our internal extension mechanism, it is helpful to begin by considering why most implementations of programming languages cannot even be  externally extended. 
-Let us consider, as a simple example, an implementation of G\"odel's T, a typed lambda calculus with recursion on primitive natural numbers  \cite{pfpl}\todo{put statics in Appendix?}. 
-A compiler for this language written using a functional language will invariably represent the primitive type families and operators of the object language using {closed} inductive datatypes. 
+Let us consider, as a simple example, an implementation of G\"odel's T, a typed lambda calculus with recursion on primitive natural numbers (see Appendix \ref{T}). 
+A compiler for this language written using a functional language will invariably represent the primitive type families and operators using {closed} inductive datatypes. 
 For example, a simple implementation in Standard ML may be based around these datatypes:
@@
   datatype Exp = Var of var 
-               | Lam of var * Type * Exp
-               | Ap of Exp * Exp 
-               | Z | S of Exp 
-               | Natrec of Exp * Exp * Exp
+               | Lam of var * Type * Exp | Ap of Exp * Exp 
+               | Z | S of Exp | Natrec of Exp * Exp * Exp
 \end{lstlisting}
 
-The logic governing typechecking and translation to an intermediate language (which suitable for optimization by later phases of compilation) will be implemented by exhaustive case analysis over the constructors of \lstinline{Exp}.
+The logic governing typechecking and translation to a suitable intermediate language (for subsequent optimization and compilation by some back-end) will proceed by exhaustive case analysis over the constructors of \lstinline{Exp}.
 
 In an object-oriented implementation of Godel's T, we might instead encode types and operators as subclasses of abstract classes \lstinline{Type} and \lstinline{Exp}.
-Typechecking and translation will proceed by the ubiquitous \emph{visitor pattern} \cite{visitor} by dispatching against a set of {known} subclasses of \lstinline{Exp}.
-
-In either case, we encounter the same basic issue: there is no way to modularly extend the representation of primitive type families and operators and implement their associated typechecking and translation logic.
-This issue is related to the widely-discussed \emph{expression problem} (we do not consider the case of adding new functions entirely here) \cite{wadler-expression}.
+Typechecking and translation will proceed by the ubiquitous \emph{visitor pattern} \cite{visitor} by dispatching against a fixed collection of {known} subclasses of \lstinline{Exp}.
+
+In either case, we encounter the same basic issue: there is no way to modularly add new primitive type families and operators and implement their associated typechecking and translation logic. 
+This issue is related to the widely-discussed \emph{expression problem} (in a restricted sense -- we do not consider adding new functions beyond typechecking and translation here, only adding logic to these) \cite{wadler-expression}.
 
@@
 In functional languages, we might use \emph{open datatypes} \cite{open-datatypes}.
-For example, if we wish to extend G\"odel's T with product types, a language supporting open datatypes might allow you to add types and operators for products to the above definitions, if they have been declared open, like this:
+For example, if we wish to extend G\"odel's T with product types and we have written our compiler in a language supporting open inductive datatypes, it might be possible to add new cases like this:
 \begin{lstlisting}
@@
 
-The logic for functionality like typechecking and translation might then be specified for only the new cases.
-For example, a function \lstinline{typeof} that assigns a type to an expression could be extended to support the \lstinline{PrL} operator like so:
+The logic for functionality like typechecking and translation could then be implemented for only these new cases.
+For example, the \lstinline{typeof} function that assigns a type to an expression could be extended like so:
 \begin{lstlisting}
-  typeof PrL(e) = 
-    case typeof e of 
+  typeof PrL(e) = case typeof e of 
       Prod(t1, _) => t1 
@@
 
-If we allow users to define new modules containing definitions like these and link them into our compiler, we have succeeded in creating an externally-extensible language implementation, albeit one where safety is not guaranteed (we will return to this point shortly).
-We have not, however, created an extensible programming language, because other implementations of the language will not necessarily support the same mechanism.
+If we allowed users to define new modules containing definitions like these and link them into our compiler, we will have succeeded in creating an externally-extensible compiler, albeit one where safety is not guaranteed (we will return to this point shortly).
+We have not, however, created an extensible programming language, for two reasons.
+First, compiler extensions are distributed and activated separately from libraries, so dependencies become more difficult to manage.
+Second, other compilers for the same language will not necessarily support the same extensions.
 If our newly-introduced constructs are exposed at a library's  interface boundary, clients using different compilers face the same problems with interoperability that those using different languages face.
-That is, \textbf{extending a language by extending a single implementation of it is equivalent to creating a new language}.
-Several prominent language ecosystems today are in a state where a prominent compiler has introduced or supported the introduction of extensions that many libraries have come to rely on, including the Glasgow Haskell Compiler, SML/NJ and the GNU compilers for C and C++.
+That is, \textbf{extending a language by extending a single compiler for it is morally equivalent to creating a new language}.
+Several prominent language ecosystems today are in a state where a prominent compiler has introduced or enabled the introduction of extensions that many libraries have come to rely on, including the Glasgow Haskell Compiler, SML/NJ and the GNU compilers for C and C++.
 We argue that this practice should be considered harmful.
 
-A more appropriate and useful place for extensions like this is directly within libraries.
-To enable this, the language must include an appropriately-constrained mechanism to introduce new type families, like \lstinline{Prod}, and operators, like \lstinline{Pair}, \lstinline{PrL} and \lstinline{PrR}, and implement the associated typechecking and translation logic.
-When encountering these new operators in expressions, the compiler must effectively  hand control over typechecking and translation to user-defined logic.
-Because this mechanism is {language-internal}, all compilers must support it to satisfy the language specification, and extensions are packaged directly within  libraries.
-
-Statically-typed languages typically make a distinction between \emph{expressions}, which describe run-time computations, and static constructs like types and datatype declarations.
+A more appropriate and useful place for extensions like this is directly within libraries, alongside abstractions that do not require this level of control.
+To enable this, the language must allow for the introduction new primitive type families, like \lstinline{Prod}, operators, like \lstinline{Pair}, \lstinline{PrL} and \lstinline{PrR}, and associated typechecking and translation logic, without allowing so much control that the reliability of the system suffers.
+When encountering these new operators in expressions, the compiler must effectively  hand control over typechecking and translation to the appropriate user-defined logic.
+Because this mechanism is {language-internal}, all compilers must support it to satisfy the language specification.
+
+Statically-typed languages typically make a distinction between \emph{expressions}, which describe run-time computations, and type-level constructs like types, type aliases and datatype declarations.
 The design described above suggests we may now need to add another layer to our language, an {extension language}, where extensions can be declared and implemented.
-In fact, we will show that \textbf{the natural place for type system extensions is within the type-level language}.
+In fact, we will show that \textbf{the most natural place for type system extensions is within the type-level language}.
 The intuition is that extensions to a language's static semantics will need to manipulate types as values at compile-time.
-Many languages already allow users to write functions over types, effectively manipulating them as values at compile-time (see Sec.
+Many languages already allow users to write type-level functions for various reasons, effectively supporting this notion of types as values at compile-time (see Sec.
 \ref{related-work} for examples).
 The type-level language is often constrained by its own type system (where the types of type-level values are called \emph{kinds} for clarity) that prevents type-level functions from causing problems during compilation.
-This is precisely the structure that a distinct extension layer would have, and so we will show that it is quite naturally to unify the two in this work.
+This is precisely the structure that a distinct extension layer would have, and so we will demonstrate that it is quite natural to unify the two in this work.
 
@@
 \textbf{programs} & \rho & ::= & \pfam{\familyDf}{\progsort} \pipe \pdef{t}{\kappa}{\tau}{\progsort} \pipe e\\
-\text{primitive ops}		&	\theta	&	::= &	\topsempty \pipe 
-												\topp{\theta}{\tops{op}{\kappai}{i}{a}{\tau}}\\
+\text{primitive ops}		&	\theta	&	::= &	\tops{op}{\kappaidx}{i}{a}{\tau} \pipe 
+												\topp{\theta}{\theta}\\
 \\
@@
 														\tnil{\kappa} \pipe \tcons{\tau_1}{\tau_2} \pipe 
-									                     \tfold{\tau_1}{\tau_2}{x}{y}{\tau_3}
+									                     \tfold{\tau_1}{\tau_2}{h}{t}{r}{\tau_3}
 														\\
@@
 
-\text{denotations} 				& 		 & 	\pipe	&	\tden{\tau_2}{\tau_1} \pipe \terr \pipe \tdencase{\tau}{y}{x}{\tau_1}{\tau_2}\\
+\text{denotations} 				& 		 & 	\pipe	&	\tden{\tauiterm}{\tautype} \pipe \terr \pipe \tdencase{\tau}{x}{t}{\tau_1}{\tau_2}\\
  %& & \pipe & 
@@
 
-\text{quoted IL}		&		&	\pipe	&	\titerm{\gamma} \pipe \titype{\sigma} \\
+\text{reified IL}		&		&	\pipe	&	\titerm{\gamma} \pipe \titype{\sigma} \\
 
@@
 												\\
- &  & \pipe & \tvalof{\tau_2}{\tau_1} \pipe \iup{\tau} \\
+ &  & \pipe & \tvalof{\tau_1}{\tau_2} \pipe \iup{\tau} \\
 %\text{deabstracted}& \gamma & ::= & \mathcal{G}[\gamma, \sigma]\\
@@
 \end{figure}
-We will now develop a core calculus, called @$\lambda$ for the ``actively-typed lambda calculus'', and discuss how to address safety concerns that arise when giving users this level of control over the semantics of a language and its implementation.
+In this section, we will develop a core calculus, called @$\lambda$ for the ``actively-typed lambda calculus'', by way of a semantics and a simple example, and discuss how it addresses the safety concerns that arise when giving users this level of control over the semantics of a language and its implementation.
 \subsection{Overview}
@@
 Kinds, $\kappa$, classify type-level terms, $\tau$.
-Types classify expressions, $e$, and are type-level values of kind $\star$ (following System $F_{\omega}$ \cite{fomega}).
+Types are type-level values of kind $\star$ (following System $F_{\omega}$ \cite{fomega}) and classify expressions, $e$.
 The type-level language also includes other kinds of terms, such as type-level functions and other type-level data structures.
@@
 \ref{termination} for design constraints in general).
+It also includes constructs for developing extensions, which we will discuss below.
 %Some kinds of type-level values (including types themselves, but excluding type-level functions) can be compared for structural equality\todo{describe this later?}.
@@
 This language has been chosen, for simplicity, to be a variant of Plotkin's PCF with primitive integers and products, but in practice would include other constructs consistent with its role as a high-level intermediate language.
-Terms of this language are denoted $\gamma$ and types are denoted $\sigma$.
-Both of these productions also include additional forms containing type-level terms, $\tau$; these are used during compilation and will be explained below.
-
-\subsection{Example: Natural Numbers}
+Terms of this language are denoted $\gamma$ and its types are denoted $\sigma$.
+Both of these productions also include additional forms containing type-level terms, $\tau$; these are used for developing extensions and during compilation and will be explained in the sections below.
+
+\subsection{Example: G\"odel's T as an Active Type Family}
 \begin{figure}[t]
@@
 \begin{flalign}
-&\family{Nat}{\kunit}{\\
-&\quad\tops{z}{\kunit}{i}{a}{
+\label{natfam}&\family{Nat}{\kunit}{\\
+\label{z}&\quad\tops{z}{\kunit}{i}{a}{
 	\tapp{\tapp{\tvar{const}}{\tvar{a}}}{
-		\tlam{\_}{\kunit}{\tden{\titerm{0}}{\ttype{Nat}{\tunit}}}
+		\tden{\titerm{0}}{\ttype{Nat}{\tunit}}
 	}};\\
-&\quad\tops{s}{\kunit}{i}{a}{
+\label{s}&\quad\tops{s}{\kunit}{i}{a}{
 	\tapp{
-		\tapp{\tvar{pop\_final}}{\tvar{a}}}{\tlam{t1}{\kTypeBlur}{\tlam{v1}{\kITerm}{
-			\\
-			&\quad\quad\tapp{\tapp{\tapp{\tvar{check\_type}}{\tvar{t1}}}{\ttype{Nat}{\tunit}}}{
-				\tden{\titerm{\iup{\tvar{v1}}+1}}{\ttype{Nat}{\tunit}}
+		\tapp{\tvar{pop\_final}}{\tvar{a}}}{\tlam{x}{\kITerm}{\tlam{t}{\kTypeBlur}{
+			\\
+			&\quad\quad\tapp{\tapp{\tapp{\tvar{check\_type}}{\tvar{t}}}{\ttype{Nat}{\tunit}}}{
+				\tden{\titerm{\iup{\tvar{x}}+1}}{\ttype{Nat}{\tunit}}
 			}
@@
 	};\\
-&\quad\tops{rec}{\kunit}{i}{a}{
-	\tapp{\tapp{\tvar{pop}}{\tvar{a}}}{\tlam{t1}{\kTypeBlur}{\tlam{v1}{\kITerm}{\tlam{a}{\klist{\kDen}}{\\
-	&\quad\quad\tapp{\tapp{\tvar{pop}}{\tvar{a}}}{\tlam{t2}{\kTypeBlur}{\tlam{v2}{\kITerm}{\tlam{a}{\klist{\kDen}}{\\
-	&\quad\quad \tapp{\tapp{\tvar{pop\_final}}{\tvar{a}}}{\tlam{t3}{\kTypeBlur}{\tlam{v3}{\kITerm}{\\
+\label{rec}&\quad\tops{rec}{\kunit}{i}{a}{
+	\tapp{\tapp{\tvar{pop}}{\tvar{a}}}{\tlam{x1}{\kITerm}{\tlam{t1}{\kTypeBlur}{\tlam{a}{\klist{\kDen}}{\\
+	&\quad\quad\tapp{\tapp{\tvar{pop}}{\tvar{a}}}{\tlam{x2}{\kITerm}{\tlam{t2}{\kTypeBlur}{\tlam{a}{\klist{\kDen}}{\\
+	&\quad\quad \tapp{\tapp{\tvar{pop\_final}}{\tvar{a}}}{\tlam{x3}{\kITerm}{\tlam{t3}{\kTypeBlur}{\\
 	&\quad\quad \tapp{\tapp{\tapp{\tvar{check\_type}}{\tvar{t1}}}{\ttype{Nat}{\tunit}}}{(\\
 	&\quad\quad \tapp{\tapp{\tapp{\tvar{check\_type}}{\tvar{t3}}}{\ttype{Arrow}{(\ttype{Nat}{\tunit},\ttype{Arrow}{(\tvar{t2},\tvar{t2})})}}}{\\
-		&\quad\quad \tden{\iapp{(\ifix{f}{\darrow{\dint}{\trepof{\tvar{t2}}}}{\ilam{x}{\dint}{\\
-		&\quad\quad\quad \iIfEq{x}{0}{\dint}{\iup{\tvar{v2}}}{\iapp{\iapp{\iup{\tvar{v3}}}{(x-1)}}{(\iapp{f}{(x-1)})}}}})}{\iup{\tvar{v1}}}}{\tvar{t2}}
+		\label{fix}&\quad\quad \tden{\titerm{\iapp{(\ifix{f}{\darrow{\dint}{\trepof{\tvar{t2}}}}{\ilam{x}{\dint}{\\
+		&\quad\quad\quad \iIfEq{x}{0}{\dint}{\iup{\tvar{x2}}}{\iapp{\iapp{\iup{\tvar{x3}}}{(x-1)}}{(\iapp{f}{(x-1)})}}}})}{\iup{\tvar{x1}}}}}{\tvar{t2}}
 	)}}
@@
 &}{XXX}{i}{\titype{\dint}};\\
-&\pdef{nat}{\kTypeBlur}{\ttype{Nat}{\tunit}}{\\
-&(\elam{plus}{\ttype{Arrow}{(\tvar{nat}, \ttype{Arrow}{(\tvar{nat}, \tvar{nat}))}}}{
+\label{nattype}&\pdef{nat}{\kTypeBlur}{\ttype{Nat}{\tunit}}{\\
+\label{plus}&(\elam{plus}{\ttype{Arrow}{(\tvar{nat}, \ttype{Arrow}{(\tvar{nat}, \tvar{nat}))}}}{
 	\elam{two}{\tvar{nat}}{\\
-	&\quad\quad \eopapp{\eopapp{plus}{two}}{two}}})
+\label{ap}	&\quad\quad \eopapp{\eopapp{plus}{two}}{two}}})
 }\\
-&\quad \elam{x}{\tvar{nat}}{\elam{y}{\tvar{nat}}{
+\label{recuse}&\quad (\elam{x}{\tvar{nat}}{\elam{y}{\tvar{nat}}{
 	\eop{Nat}{rec}{\tunit}{x; y; \elam{p}{\tvar{nat}}{\elam{r}{\tvar{nat}}{
 	\eop{Nat}{s}{\tunit}{r}
-	}}}}}\\
-&\quad  \eop{Nat}{s}{\tunit}{\eop{Nat}{s}{\tunit}{\eop{Nat}{z}{\tunit}{ }}}
+	}}}}})\\
+\label{two}&\quad  \eop{Nat}{s}{\tunit}{\eop{Nat}{s}{\tunit}{\eop{Nat}{z}{\tunit}{ }}}
 %{\\&\elet{two}{\tvar{nat}}{\eop{Nat}{s}{\tunit}{\eop{Nat}{s}{\tunit}{\eop{Nat}{z}{\tunit}{ }}}}{\eapp{\eapp{plus}{two}}{two}}}
@@
 \end{flalign}
-\caption{Example: primitive natural numbers in \atlam. Type-level helper functions $\tvar{const}$, $\tvar{pop}$, $\tvar{pop\_final}$ and $\tvar{check\_type}$ are given in Appendix \ref{helper}.}
+\vspace{-15pt}
+\caption{G\"odel's T in \atlam, used to calculate 2+2.
+The statics we implement here are shown in Appendix \ref{T}.
+Simple helper functions for working with argument lists ($\tvar{const}$, $\tvar{pop}$, $\tvar{pop\_final}$) and types ($\tvar{check\_type}$) are defined in Appendix \ref{helper}.}
 \label{example}
 \end{figure}
-To make our discussion of each of these components of the calculus concrete, we will begin with an example showing how to introduce primitive natural numbers, implemented as integers internally, as a user-defined extension.
-
-\subsection{Central Compilation Judgement}
+To make our explanation of each of the constructs in the calculus concrete, we will work through an example showing how to introduce primitive natural numbers with bounded recursion in the style of G\"odel's T \cite{pfpl}.
+These will be implemented internally as integers (that is, translation will always produce internal terms of internal type $\dint$).
+Figure \ref{example} shows how the indexed type family $\fvar{Nat}$ is defined.
+This family is indexed by a unit value (of kind $\kunit$), so it contains only one type, $\ttype{Nat}{\tunit}$, which we alias by defining the type-level variable $\tvar{nat}$.
+We define the typechecking and translation logic for the operators associated with this family (\opvar{z}, \opvar{s} and \opvar{rec}) and use these to define an add function and compute 2+2.
+
+\subsection{Indexed Type Families and Types}\label{families}
+The syntactic form $\familyDf$ declares a new primitive type family named $\fvar{Fam}$ indexed by type-level values of kind $\kappaidx$ (line \ref{natfam}).
+This can be compared to adding a new constructor to the compiler-internal datatype \lstinline{Type}, as suggested in Sec.
+\ref{evolution}.
+The index represents the data associated with the constructor.
+A base type like $\tvar{nat}$ can be thought of as being the only type in the family $\fvar{Nat}$ trivially indexed by the unit value, of kind $\kunit$, while families like $\fvar{Ntuple}$ might be indexed by a list of types, of kind $\klist{\kTypeBlur}$.
+A type (that is, a type-level term of kind $\kTypeBlur$) is constructed by naming a family in scope and providing a type-level term of the appropriate kind as an index.
+For example, $\ttype{Ntuple}{\tcons{\tvar{nat}}{\tcons{\tvar{nat}}{\tnil{\kTypeBlur}}}}$ might be the type of a pair of natural numbers.
+It is important that type equality be decidable, so only kinds for which equivalence coincides with syntactic equality can be used as type family indices.
+The main  consequence of this restriction is that indices cannot contain type-level functions.
+
+As we will discuss further below, it is important that all expressions classified by a type compile to internal terms of consistent internal type\todo{mention compiler correctness history here?}.
+For this reason, we require that every type have an associated internal type.
+This is computed by substituting the type index for the bound variable $\tvar{i}$ in the term $\taurep$, called the \emph{representation schema}, associated with the type family and reducing to a value representing an internal type.
+Internal types, $\sigma$, are reified as type-level terms of kind $\kIType$ using the introductory form $\titype{\sigma}$ (as we will see, an  elimination form  is not necessary).
+
+\subsection{Indexed Operator Families and Denotations}\label{operators}
+Type families also have a collection of primitive operator families associated with them.
+An operator family named $\opvar{op}$ is declared using the form $\tops{op}{\kappaidx}{i}{a}{\tauop}$.
+Like type families, operator families are indexed by values of some specified kind, $\kappaidx$.
+Operators are not first-class type-level values in our calculus, so there are no restrictions related to equality.
+In the example in Fig.
+\ref{example}, all the operators are trivially indexed by the kind $\kunit$, so each family is only inhabited by one operator.
+However, a family like $\fvar{Ntuple}$ would have a family of projection operators, $\opvar{pr}$, indexed by a position (e.g.
+an integer in our calculus).
+A family implementing record types or object types might have a similar operator indexed by a type-level string representing the field being accessed.
+
+An operator application is an expression of the form $\eop{Fam}{op}{\tauidx}{e_1; \ldots; e_n}$, where $n \geq 0$.
+The typechecking and translation of an expression of this form is controlled by the declaration of $\opvar{op}$ in family $\fvar{Fam}$, as determined by the type-level term $\tauop$.
+It must evaluate to a \emph{denotation}, which is a type-level value of kind $\kDen$, when the index, $\tauidx$, and a list containing the denotations recursively assigned to each argument, $e_1$ through $e_n$, are bound to $\tvar{i}$ and $\tvar{a}$, respectively.
+
+There are two forms of denotations.
+A \emph{valid denotation} has the form $\tden{\tau_1}{\tau_2}$, where $\tau_1$ represents the \emph{translation} of an expression to an internal term and $\tau_2$ is the type.
+The translation must have kind $\kITerm$ of  reified internal language terms.
+The only introductory form for this is $\titerm{\gamma}$, just as with $\kIType$.
+There is no elimination form.
+Types, as before, have kind $\kTypeBlur$.
+%(indeed, in \atlam, terms are never  syntactically analyzed directly by extensions, unlike macro systems and other forms of term rewriting; see Sec. \ref{related-work})
+If a type cannot be assigned to an operation according to the specification of the operator being implemented, then the \emph{error denotation}, $\terr$, should be returned instead of a valid denotation.
+In a practical implementation, a specialized error message and other diagnostic information could be associated with $\terr$, but we omit such details for the sake of simplicity in our core calculus.
+Terms of kind $\kDen$ can be \textsf{case} analyzed to determine if they are valid, and if so, to extract the translation and type.
+
+In the example in Fig.
+\ref{example}, the operator $\opvar{z}$ checks that no arguments were provided using the simple helper function $\tvar{const}$ (shown in Appendix \ref{helper}).
+If so, it returns a valid denotation by pairing the translation $\titerm{0}$ with the type $\ttype{Nat}{\tunit}$, as expected\footnote{In fact, there is no theoretical barrier to a different ``zero'' being used, since integers and natural numbers are both of countably infinite cardinality!}.
+If not, the helper function returns $\terr$.
+The successor operator, $\opvar{s}$, does take an argument, so it pops a denotation off the argument list, making sure there are no more, and binds its translation and type to $\tvar{x}$ and $\tvar{t}$ respectively, all using the helper function $\tvar{pop\_final}$.
+It then checks that the argument's type is also $\ttype{Nat}{\tunit}$, returning a denotation pairing the translation $\titerm{\iup{\tvar{x}} + 1}$ with the type $\ttype{Nat}{\tunit}$ if so.
+The internal language includes a special form $\iup{\tau}$ which is used to ``un-reify'' reified internal terms, of kind $\kITerm$ (thus serving as the left-inverse of $\titerm{\gamma}$).
+In this case, $\tvar{x}$ is a type-level variable of kind $\kITerm$ representing the translation of the argument to the successor operator.
+Because we have checked that it is a natural number, and natural numbers always translate to integers by the representation schema, we know that it is safe to add 1 to this internal term.
+The result remains an integer according to the type system of the internal language so it is safe to give the entire expression type $\ttype{Nat}{\tunit}$ without violating the schema.
+Indeed, the extension mechanism will check that these consistencies hold (though not the kind system; see Sec.
+\ref{repcon}).
+If any of these steps fail, the various helper functions we use simply return $\terr$ (in practice, it would be prudent to equip each failure condition with a different error message).
+
+% The operator index, $\tauidx$, and .
+
+\subsection{Functions, Variables and Arrow Types}
+The recursor operator, $\opvar{rec}$, proceeds similarly, extracting the translations and types of each of its three arguments.
+The third argument, the recursive case, binds two variables (the predecessor and the result of recursing on it).
+In \atlam, the build-in $\lambda$ operator serves as the primitive mechanism for working with bound variables.
+While lambda terms conventionally have types of the form $\tau_1 \rightarrow \tau_2$, \atlam~instead includes a built-in type family, $\fvar{Arrow}$,  that is always in scope.
+This family is indexed by a pair of types, $\kpair{\kTypeBlur}{\kTypeBlur}$, and its representation schema simply maps to the corresponding internal arrow type, $\tvar{i}.\titype{\darrow{\trepof{\tfst{\tvar{i}}}}{\trepof{\tsnd{\tvar{i}}}}}$.
+The special form $\trepof{\tau}$ is used to refer, abstractly, to the representation of another type (we also see this used on line \ref{fix}).
+Although $\lambda$ is built-in, application is an operator associated with the $\fvar{Arrow}$ family, $\opvar{ap}$, as we see on line \ref{ap}.
+\todo{figure for arrow types}
+
+\subsection{Compilation}
 \begin{figure}[t]
@@
 The \emph{central compilation judgement}, shown in Figure \ref{ccj}, captures the two phases of compilation: kind checking and active typechecking and translation. 
-The latter phase requires normalizing type-level terms to type-level values, so the  kind checking phase preceding it will ensure that this process not ``get stuck'', as we will elaborate upon when discussing safety and decidability in Sec.
-\ref{safety}.
-
-\subsection{Indexed Type Families and Types}\label{families}
-Declaring a new primitive indexed type family by using \textsf{family} can be compared to adding a new constructor to the compiler's \lstinline{Type} datatype, as suggested in Section \ref{evolution}.
-The kind of type-level data that the family is indexed by is specified by $\kappaidx$.
-For example, a base type like $\tvar{nat}$ can be thought of as being the only type in a family, $\fvar{Nat}$, trivially indexed by the unit value, of kind $\kunit$, while families like $\fvar{Pair}$ might be indexed by a pair of types, of kind $\kpair{\kTypeBlur}{\kTypeBlur}$.
-A type (that is, a type-level term of kind $\kTypeBlur$) is created by naming a family that has been previously been introduced and providing an index of the appropriate kind.
-For example, $\ttype{Prod}{\tpair{\tvar{nat}}{\tvar{nat}}}$ is the type of a pair of natural numbers in a context where the families $\fvar{Prod}$ and $\fvar{Nat}$ have been introduced as described above and the type-level variable $\tvar{nat}$ has been bound to the type $\ttype{Nat}{\tunit}$.
-
-\subsection{Indexed Operator Families and Active Types}\label{operators}
-We build in lambdas as the only binding structure in the language to simplify the core calculus. All other operators are associated with a type. For example, nat / tuples / records.
-
-Operator families.
+The first phase ensures that all type-level terms in the program are well-kinded and that all expressions and internal terms are closed.
+The kinding rules for programs are given in Figure \ref{kindprog}, and they rely on the kinding rules for type-level terms given in Figure \ref{tlkind}.
+These rules use contexts $\Sigma$ and $\Theta$ to track family and operator signatures but beyond that, these rules are largely consistent with those of a simply-typed lambda calculus, with the addition of the handful of special forms described in the previous sections.
+The reader is encouraged to verify that the example in Fig.
+\ref{example} is well-kinded.
+Active typechecking and translation requires normalizing type-level terms to type-level values at various points, so the kind checking phase ensures that no simple mistakes have been made and that this process will not ``get stuck'' (see Sec.
+\ref{safety}).
 
@@
 \subsection{Abstract Representations}
-
+Note that there is no elimination form that could be used to determine the internal type associated with a type, but the special internal type $\trepof{\tau}$, where $\tau$ is a type, can be used to refer to it abstractly
 \begin{figure}[t]
@@
 \begin{mathpar}
-\inferrule[family decl kinding]{
-	\fvar{Fam} \notin \fvarCtx\\
+\inferrule[family kinding]{
+	\fvar{Fam} \notin \text{dom}(\fvarCtx)\\
 	\kEq{\kappaidx}\\
@@
 \begin{mathpar}
-\inferrule[no ops]{ }{
-	\tKindX{\topsempty}{\kOpEmpty}
+\inferrule[op]{
+	\tKind{\tvarCtx, \tOfKind{\tvar{i}}{\kappai}, \tOfKind{\tvar{a}}{\klist{\kDen}}}{\fvarCtx}{\tau}{\kDen}
+}{
+	\opType{\tvarCtx}{\fvarCtx}{{\tops{op}{\kappai}{i}{a}{\tau}}}{
+	\kOpS{op}{\kappai}}
 }
@@
 \inferrule[ops]{
-	\opType{\tvarCtx}{\fvarCtx}{\theta}{\Theta}\\
-	\opvar{op} \notin \theta\\
-	\tKind{\tvarCtx, \tOfKind{\tvar{i}}{\kappai}, \tOfKind{\tvar{a}}{\klist{\kDen}}}{\fvarCtx}{\tau}{\kDen}
-}{
-	\opType{\tvarCtx}{\fvarCtx}{\topp{\theta}{\tops{op}{\kappai}{i}{a}{\tau}}}{
-	\kOp{\Theta}{op}{\kappai}}
+	\tKindX{\theta_1}{\Theta_1}\\
+	\tKindX{\theta_2}{\Theta_2}\\\\
+	\text{dom}(\theta_1) \cap \text{dom}(\theta_2) = \emptyset
+}{
+	\tKindX{\theta_1; \theta_2}{\Theta_1, \Theta_2}
 }
@@
 \end{mathpar}
-\caption{\small Kinding for programs. Variable contexts $\tvarCtx$ and $\itvarCtx$ obey standard structural properties. Kinding for type-level terms in Figure \ref{kindtl}. \ \label{kindprof}}
+\caption{\small Kinding for programs. Variable contexts $\tvarCtx$ and $\itvarCtx$ obey standard structural properties. Kinding for type-level terms in Figure \ref{kindtl}.}
+\label{kindprog}
 \end{figure}
@@
 \caption{\small Kinding for type-level terms}
+\label{tlkind}
 \end{figure}
@@
 \caption{\small Evaluation semantics for type-level terms}
+\label{tleval}
 \end{figure}
@@
 \caption{\small Abstracted internal typing}
+\label{ait}
 \end{figure}
@@
 \caption{\small Deabstraction rules}
+\label{deabs}
 \end{figure}
@@
 \section{Appendix}
+\subsection{Statics of G\"odel's T}\label{T}
+\newcommand{\lam}[3]{\lambda #1{:}#2.#3}
+\newcommand{\ap}[2]{#1~#2}
+\newcommand{\z}{\textsf{z}}
+\newcommand{\s}[1]{\textsf{s}(#1)}
+\newcommand{\natrec}[5]{\textsf{natrec}(#1; #2; #3,#4.#5)}
+\newcommand{\pair}[2]{(#1, #2)}
+\newcommand{\fst}[1]{\textsf{fst}(#1)}
+\newcommand{\snd}[1]{\textsf{snd}(#1)}
+
+\newcommand{\tArrow}[2]{#1 \rightarrow #2}
+\newcommand{\nat}{\texttt{nat}}
+\renewcommand{\prod}[2]{#1 \times #2}
+
+%\newcommand{\eCtx}{\Gamma}
+\newcommand{\eCtxX}[3]{#1, #2 : #3}
+\newcommand{\jet}[3]{#1 \vdash #2 : #3}
+\newcommand{\jetX}[2]{\jet{\eCtx}{#1}{#2}}
+%\begin{figure}
+%\small
+\begin{mathpar}
+\small
+\inferrule[var]{ }{
+	\jet{\eCtxX{\eCtx}{x}{\tau}}{x}{\tau}
+}
+
+\inferrule[Arrow-I]{
+	\jet{\eCtxX{\eCtx}{x}{\tau}}{e}{\tau'}	
+}{
+	\jetX{\lam{x}{\tau}{e}}{\tArrow{\tau}{\tau'}}
+}
+
+\inferrule[Arrow-E]{
+	\jetX{e_1}{\tArrow{\tau}{\tau'}}\\
+	\jetX{e_2}{\tau}
+}{
+	\jetX{\ap{e_1}{e_2}}{\tau'}
+}
+
+\inferrule[Nat-I1]{ }{
+	\jetX{\z}{\nat}
+}
+~~~~~~~~
+\inferrule[Nat-I2]{
+	\jetX{e}{\nat}
+}{
+	\jetX{\s{e}}{\nat}
+}
+~~~~~~~~
+\inferrule[Nat-E]{
+	\jetX{e_1}{\nat}\\
+	\jetX{e_2}{\tau}\\
+	\jetX{e_3}{\tArrow{\nat}{\tArrow{\tau}{\tau}}}
+}{
+	\jetX{\natrec{e_1}{e_2}{x}{y}{e_3}}{\tau}
+}
+\end{mathpar}
+%\caption{Static semantics of G\"odel's T}
+%\end{figure}
+% \begin{figure}
+% \small
+% \begin{mathpar}
+% \inferrule[Prod-I]{
+% 	\jetX{e_1}{\tau_1}\\
+% 	\jetX{e_2}{\tau_2}
+% }{
+% 	\jetX{\pair{e_1}{e_2}}{\prod{\tau_1}{\tau_2}}
+% }
+
+% \inferrule[Prod-E1]{
+% 	\jetX{e}{\prod{\tau_1}{\tau_2}}
+% }{
+% 	\jetX{\fst{e}}{\tau_1}
+% }
+
+% \inferrule[Prod-E2]{
+% 	\jetX{e}{\prod{\tau_1}{\tau_2}}
+% }{
+% 	\jetX{\snd{e}}{\tau_2}
+% }
+% \end{mathpar}
+% \caption{Static semantics of products}\label{prods}
+% \end{figure}
 \subsection{Helper Functions}\label{helper}
-
+\todo{describe these}
+\[\begin{array}{rcl}
+\tvar{const} & := & (\tlam{a}{\klist{\kDen}}{\tlam{d}{\kDen}{\tfold{\tvar{a}}{\tvar{d}}{\_}{\_}{\_}{\terr}}})\\
+\tvar{pop} & := & (\tlam{a}{\klist{\kDen}}{\tlam{f}{\karrow{\kITerm}{\karrow{\kTypeBlur}{\karrow{\klist{\kDen}}{\kDen}}}}{\\& & \quad\tfold{\tvar{a}}{\terr}{d}{b}{\_}{\tdencase{\tvar{d}}{x}{t}{
+	\tapp{\tapp{\tapp{\tvar{f}}{\tvar{x}}}{\tvar{t}}}{\tvar{b}}}{\terr}}}})\\
+\tvar{pop\_final} & := & (\tlam{a}{\klist{\kDen}}{\tlam{f}{\karrow{\kITerm}{\karrow{\kTypeBlur}{\kDen}}}{\\& & \quad\tfold{\tvar{a}}{\terr}{d}{b}{\_}{\\& & \quad \tfold{\tvar{b}}{\tdencase{\tvar{d}}{x}{t}{\tapp{\tapp{\tvar{f}}{\tvar{x}}}{\tvar{t}}}{\terr}}{\_}{\_}{\_}{\terr}}}})\\
+\tvar{check\_type} & := & (\tlam{t1}{\kTypeBlur}{\tlam{t2}{\kTypeBlur}{\tlam{d}{\kDen}{\tifeq{\tvar{t1}}{\tvar{t2}}{\kTypeBlur}{\tvar{d}}{\terr}}}})
+\end{array}\]
 \end{document}

### 2013-10-12 esop14/att-esop14.tex
Up through abs reps

@@
 For example, regular expressions encoded using inductive datatypes are often considered overly verbose, so most functional languages support them via strings, which is less safe.
-Finally, general-purpose abstractions are also implemented in a uniform manner.
-Domain knowledge is thus not applied to eliminate overhead or perform optimizations, and implementations designed for typical application workloads may not be satisfactory in parts of a program where performance is a key criteria, particularly when targeting heterogeneous hardware platforms (e.g.
+Finally, general-purpose abstractions are implemented in a uniform manner.
+Domain knowledge is not easily applied to eliminate overhead or perform optimizations, and implementations designed for typical application workloads may not be satisfactory in parts of a program where performance is a key criteria, particularly when targeting heterogeneous hardware platforms (e.g.
 programmable GPUs) and distributed computing resources.
+Using variants of familiar abstractions backed by specialized implementation strategies are quite useful in these cases.
 
@@
 We diagram this fundamental \emph{interoperability problem} in Figure \ref{approaches}(a).
-As an example, F\#'s type system prevents \lstinline{null} values from occurring within data structures, but because it's type system is not available when calling into F\# code from another language, like C\#, run-time null checks must still be inserted.
+As an example, F\#'s type system prevents \lstinline{null} values from occurring within data structures, but because it's type system is not available when calling into F\# code from another language, like C\#, run-time null checks must still be included in the implementation.
 \begin{figure}[t]
@@
 
-\emph{Internally-extensible programming languages} promise to avoid these problems by providing researchers and domain experts with a mechanism for implementing the semantics of new primitive constructs directly within libraries\todo{mention active libraries}.
+\emph{Internally-extensible programming languages} promise to avoid these problems by providing researchers and domain experts with a mechanism for implementing the semantics of new primitive constructs directly within libraries.
 %Developers of libraries need only determine whether they are appropriate for their domain, without also considering whether these constructs can be exposed in terms of abstractions appropriate to client code.
@@
 Providers of components thus need only consider whether primitives that they use are appropriate for their domain, without also considering whether their code might be used in a context where these primitives are not otherwise appropriate.
-Libraries containing logic that is invoked at compile-time have been called \emph{active libraries} \cite{activelibraries}.
-We diagram this competing approach in Figure \ref{approaches}(b).
+Libraries containing logic that is invoked at compile-time, as extension logic would be, have been called \emph{active libraries} \cite{activelibraries}.
+We adopt this terminology and diagram this competing approach in Figure \ref{approaches}(b).
 
@@
 \subsection{Overview}
-\todo{remove $\Theta$ from family}The grammar of \atlam~is shown in Figure \ref{grammar}.
-The language is a simply-typed lambda calculus with simply-kinded type-level computation.
-Kinds, $\kappa$, classify type-level terms, $\tau$.
+The grammar of \atlam~is shown in Figure \ref{grammar}. The language is a simply-typed lambda calculus with simply-kinded type-level computation. Kinds, $\kappa$, classify type-level terms, $\tau$. 
 Types are type-level values of kind $\star$ (following System $F_{\omega}$ \cite{fomega}) and classify expressions, $e$.
-The type-level language also includes other kinds of terms, such as type-level functions and other type-level data structures.
-We include lists (required by our mechanism) as well as type-level integers, strings and products for the sake of our examples (see Sec.
-\ref{termination} for design constraints in general).
-It also includes constructs for developing extensions, which we will discuss below.
-%Some kinds of type-level values (including types themselves, but excluding type-level functions) can be compared for structural equality\todo{describe this later?}.
+The type-level language also includes other kinds of terms, such as type-level functions, lists (required by our mechanism) and integers, strings and products for the sake of our examples (see Sec.
+\ref{termination} for a discussion on other acceptable kinds of type-level data).
+It also includes constructs for developing extensions -- denotations and reified internal terms and types -- which we will discuss in the sections below.
 
 At the top level, programs, $\rho$, consist of a series of declarations followed by an expression.
-Declarations can be either bindings of type-level terms to type-level variables using \textsf{def} or a declaration of a new primitive type family using \textsf{family} (Sec.
-\ref{families}).
-Expressions can be either variables, lambdas, or applications of operators (Sec.
-\ref{operators}), and are ultimately given meaning by translation to a typed internal language.
+Declarations can be either bindings of type-level terms to type-level variables using \textsf{def} or a declaration of a new primitive type family using \textsf{family}.
+Expressions can be either variables, lambdas, or applications of operators, and are ultimately given meaning by translation to a typed internal language.
 This language has been chosen, for simplicity, to be a variant of Plotkin's PCF with primitive integers and products, but in practice would include other constructs consistent with its role as a high-level intermediate language.
-Terms of this language are denoted $\gamma$ and its types are denoted $\sigma$.
-Both of these productions also include additional forms containing type-level terms, $\tau$; these are used for developing extensions and during compilation and will be explained in the sections below.
+The grammars of internal terms, $\gamma$, and internal types, $\sigma$, also include special forms containing type-level terms; these are used for developing extensions and during compilation and will be erased before compilation ends, as we will explain below.
 
@@
 		\label{fix}&\quad\quad \tden{\titerm{\iapp{(\ifix{f}{\darrow{\dint}{\trepof{\tvar{t2}}}}{\ilam{x}{\dint}{\\
-		&\quad\quad\quad \iIfEq{x}{0}{\dint}{\iup{\tvar{x2}}}{\iapp{\iapp{\iup{\tvar{x3}}}{(x-1)}}{(\iapp{f}{(x-1)})}}}})}{\iup{\tvar{x1}}}}}{\tvar{t2}}
+		\label{lastop}&\quad\quad\quad \iIfEq{x}{0}{\dint}{\iup{\tvar{x2}}}{\iapp{\iapp{\iup{\tvar{x3}}}{(x-1)}}{(\iapp{f}{(x-1)})}}}})}{\iup{\tvar{x1}}}}}{\tvar{t2}}
 	)}}
@@
 }\\
-\label{recuse}&\quad (\elam{x}{\tvar{nat}}{\elam{y}{\tvar{nat}}{
+\label{add}&\quad (\elam{x}{\tvar{nat}}{\elam{y}{\tvar{nat}}{
 	\eop{Nat}{rec}{\tunit}{x; y; \elam{p}{\tvar{nat}}{\elam{r}{\tvar{nat}}{
@@
 \label{example}
+\vspace{-10pt}
 \end{figure}
 To make our explanation of each of the constructs in the calculus concrete, we will work through an example showing how to introduce primitive natural numbers with bounded recursion in the style of G\"odel's T \cite{pfpl}.
-These will be implemented internally as integers (that is, translation will always produce internal terms of internal type $\dint$).
+These will be implemented internally as integers (internal terms of internal type $\dint$).
 Figure \ref{example} shows how the indexed type family $\fvar{Nat}$ is defined.
-This family is indexed by a unit value (of kind $\kunit$), so it contains only one type, $\ttype{Nat}{\tunit}$, which we alias by defining the type-level variable $\tvar{nat}$.
-We define the typechecking and translation logic for the operators associated with this family (\opvar{z}, \opvar{s} and \opvar{rec}) and use these to define an add function and compute 2+2.
+This family is indexed by a unit value (of kind $\kunit$), so it contains only one type, $\ttype{Nat}{\tunit}$, which we alias on line \ref{nattype} by defining the type-level variable $\tvar{nat}$.
+We define the typechecking and translation logic for the operators associated with this family (\opvar{z}, \opvar{s} and \opvar{rec}) on lines \ref{z}-\ref{lastop} and use these to define a $plus$ function on line \ref{add} and compute 2+2.
 
 \subsection{Indexed Type Families and Types}\label{families}
-The syntactic form $\familyDf$ declares a new primitive type family named $\fvar{Fam}$ indexed by type-level values of kind $\kappaidx$ (line \ref{natfam}).
+The syntactic form $\familyDf$ declares a new primitive type family named $\fvar{Fam}$ indexed by type-level values of kind $\kappaidx$ with representation schema $\tvar{i}.\taurep$ and operators $\theta$.
 This can be compared to adding a new constructor to the compiler-internal datatype \lstinline{Type}, as suggested in Sec.
@@
 The index represents the data associated with the constructor.
-A base type like $\tvar{nat}$ can be thought of as being the only type in the family $\fvar{Nat}$ trivially indexed by the unit value, of kind $\kunit$, while families like $\fvar{Ntuple}$ might be indexed by a list of types, of kind $\klist{\kTypeBlur}$.
+A base type like $\tvar{nat}$ can be thought of as being the only type in the family $\fvar{Nat}$ trivially indexed by the unit value, of kind $\kunit$, while families like $\fvar{Ntuple}$ might be indexed by a list of types, having kind $\klist{\kTypeBlur}$.
 A type (that is, a type-level term of kind $\kTypeBlur$) is constructed by naming a family in scope and providing a type-level term of the appropriate kind as an index.
@@
 The main  consequence of this restriction is that indices cannot contain type-level functions.
-
-As we will discuss further below, it is important that all expressions classified by a type compile to internal terms of consistent internal type\todo{mention compiler correctness history here?}.
-For this reason, we require that every type have an associated internal type.
-This is computed by substituting the type index for the bound variable $\tvar{i}$ in the term $\taurep$, called the \emph{representation schema}, associated with the type family and reducing to a value representing an internal type.
-Internal types, $\sigma$, are reified as type-level terms of kind $\kIType$ using the introductory form $\titype{\sigma}$ (as we will see, an  elimination form  is not necessary).
+Given a type, its family can be \textsf{case} analyzed to extract its index.
+
+\subsection{Representation Schemas}
+As we will discuss further below, it is important that all expressions classified by a type compile to consistently-typed internal terms\todo{mention compiler correctness history here?}.
+For this reason, we require that every type have associated with it a single internal type.
+This is computed by substituting the type index for the bound variable $\tvar{i}$ in the term $\taurep$, called the \emph{representation schema}, associated with the type family and evaluating to a value representing an internal type.
+Internal types, $\sigma$, are reified into type-level terms of kind $\kIType$ using the introductory form $\titype{\sigma}$.
 
@@
 An operator family named $\opvar{op}$ is declared using the form $\tops{op}{\kappaidx}{i}{a}{\tauop}$.
-Like type families, operator families are indexed by values of some specified kind, $\kappaidx$.
-Operators are not first-class type-level values in our calculus, so there are no restrictions related to equality.
+Like type families, operator families are indexed by values of some kind, $\kappaidx$.
+Operators are not first-class type-level values in our calculus, so there are no restrictions on this kind related to equality.
 In the example in Fig.
-\ref{example}, all the operators are trivially indexed by the kind $\kunit$, so each family is only inhabited by one operator.
-However, a family like $\fvar{Ntuple}$ would have a family of projection operators, $\opvar{pr}$, indexed by a position (e.g.
+\ref{example}, all the operators are trivially indexed by the kind $\kunit$, so each family only contains one operator.
+However, a type family like $\fvar{Ntuple}$ would be equipped with a family of projection operators, $\opvar{pr}$, indexed by a position (e.g.
 an integer in our calculus).
@@
 
-An operator application is an expression of the form $\eop{Fam}{op}{\tauidx}{e_1; \ldots; e_n}$, where $n \geq 0$.
-The typechecking and translation of an expression of this form is controlled by the declaration of $\opvar{op}$ in family $\fvar{Fam}$, as determined by the type-level term $\tauop$.
-It must evaluate to a \emph{denotation}, which is a type-level value of kind $\kDen$, when the index, $\tauidx$, and a list containing the denotations recursively assigned to each argument, $e_1$ through $e_n$, are bound to $\tvar{i}$ and $\tvar{a}$, respectively.
-
-There are two forms of denotations.
-A \emph{valid denotation} has the form $\tden{\tau_1}{\tau_2}$, where $\tau_1$ represents the \emph{translation} of an expression to an internal term and $\tau_2$ is the type.
-The translation must have kind $\kITerm$ of  reified internal language terms.
-The only introductory form for this is $\titerm{\gamma}$, just as with $\kIType$.
-There is no elimination form.
-Types, as before, have kind $\kTypeBlur$.
+To select an operator from a family by providing an index and apply it to $n$ arguments, the grammar provides a uniform form: $\eop{Fam}{op}{\tauidx}{e_1; \ldots; e_n}$, where $n \geq 0$.
+The typechecking and translation of an expression of this form is controlled by the implementation, $\tauop$.
+It must evaluate to a \emph{denotation}, which is a type-level value of kind $\kDen$, when given the operator index, $\tauidx$, and a list constructed from the denotations recursively assigned to each argument, $e_1$ through $e_n$.
+There are two forms of denotations that an expression can have.
+A \emph{valid denotation} has the form $\tden{\tauiterm}{\tautype}$, where $\tauiterm$ represents the \emph{translation} of the expression to an internal term and $\tautype$ is the type assigned to it.
+Internal terms are reified as type-level terms using the form $\titerm{\gamma}$, and have kind $\kITerm$ (similar to $\kIType$).
+It is important to note that there are no elimination form for reified terms or types.
 %(indeed, in \atlam, terms are never  syntactically analyzed directly by extensions, unlike macro systems and other forms of term rewriting; see Sec. \ref{related-work})
-If a type cannot be assigned to an operation according to the specification of the operator being implemented, then the \emph{error denotation}, $\terr$, should be returned instead of a valid denotation.
-In a practical implementation, a specialized error message and other diagnostic information could be associated with $\terr$, but we omit such details for the sake of simplicity in our core calculus.
-Terms of kind $\kDen$ can be \textsf{case} analyzed to determine if they are valid, and if so, to extract the translation and type.
+If a type cannot be assigned to an operation according to the specification of the operator being implemented, then the \emph{error denotation}, $\terr$, is returned instead of a valid denotation.
+In a more practical implementation, a specialized error message and other diagnostic information could be associated with $\terr$, but we omit such details for the sake of simplicity in our core calculus.
+Terms of kind $\kDen$ can be \textsf{case} analyzed to determine if they are valid or errors, and if valid, to extract the translation and type.
 
@@
 The recursor operator, $\opvar{rec}$, proceeds similarly, extracting the translations and types of each of its three arguments.
-The third argument, the recursive case, binds two variables (the predecessor and the result of recursing on it).
-In \atlam, the build-in $\lambda$ operator serves as the primitive mechanism for working with bound variables.
-While lambda terms conventionally have types of the form $\tau_1 \rightarrow \tau_2$, \atlam~instead includes a built-in type family, $\fvar{Arrow}$,  that is always in scope.
-This family is indexed by a pair of types, $\kpair{\kTypeBlur}{\kTypeBlur}$, and its representation schema simply maps to the corresponding internal arrow type, $\tvar{i}.\titype{\darrow{\trepof{\tfst{\tvar{i}}}}{\trepof{\tsnd{\tvar{i}}}}}$.
-The special form $\trepof{\tau}$ is used to refer, abstractly, to the representation of another type (we also see this used on line \ref{fix}).
-Although $\lambda$ is built-in, application is an operator associated with the $\fvar{Arrow}$ family, $\opvar{ap}$, as we see on line \ref{ap}.
+The interesting part comes when analyzing the third argument, the recursive case, which binds two variables (the predecessor and the result of recursing on it).
+In \atlam, the built-in $\lambda$ operator serves as the sole mechanism for introducing bound variables.
+In most calculi, lambda terms have types of the form $\tau_1 \rightarrow \tau_2$, which in \atlam~corresponds to a built-in type family, $\fvar{Arrow}$, indexed by a pair of types, $\kpair{\kTypeBlur}{\kTypeBlur}$, that is always in scope.
+Its representation schema simply maps to the corresponding internal arrow type, $\tvar{i}.\titype{\darrow{\trepof{\tfst{\tvar{i}}}}{\trepof{\tsnd{\tvar{i}}}}}$.
+The special form $\trepof{\tau}$ is used to refer, abstractly, to the representation of the two types it is indexed by (we also see this used on line \ref{fix}).
+Although $\lambda$ is built-in, application is an operator associated with the $\fvar{Arrow}$ family, $\opvar{ap}$, as is seen on line \ref{ap}.
 \todo{figure for arrow types}
@@
 The kinding rules for programs are given in Figure \ref{kindprog}, and they rely on the kinding rules for type-level terms given in Figure \ref{tlkind}.
-These rules use contexts $\Sigma$ and $\Theta$ to track family and operator signatures but beyond that, these rules are largely consistent with those of a simply-typed lambda calculus, with the addition of the handful of special forms described in the previous sections.
+These rules use contexts $\Sigma$ and $\Theta$ to track family and operator signatures, but beyond that, the rules are largely consistent with those of a simply-typed lambda calculus, with the addition of the handful of special forms constrained as described in the previous sections.
 The reader is encouraged to verify that the example in Fig.
 \ref{example} is well-kinded.
-Active typechecking and translation requires normalizing type-level terms to type-level values at various points, so the kind checking phase ensures that no simple mistakes have been made and that this process will not ``get stuck'' (see Sec.
+% As we will see, the kind checking phase ensures that many kinds of errors are ruled out and that this process will not ``get stuck'' (see Sec.
 \ref{safety}).
-
-\subsection{Active Typechecking and Translation}
-
-\subsection{Abstract Representations}
-Note that there is no elimination form that could be used to determine the internal type associated with a type, but the special internal type $\trepof{\tau}$, where $\tau$ is a type, can be used to refer to it abstractly
+The evaluation semantics for type-level terms are given in Fig.
+\ref{tleval}.
 \begin{figure}[t]
@@
 \begin{mathpar}
-\inferrule[family kinding]{
+\inferrule[family-kinding]{
 	\fvar{Fam} \notin \text{dom}(\fvarCtx)\\
@@
 
-\inferrule[type-level bind kinding]{
+\inferrule[def-kinding]{
 	\tKindX{\tau}{\kappa}\\
@@
 
-\inferrule[exp kinding]{
+\inferrule[exp-kinding]{
 	\exprOK{\tvarCtx}{\emptyctx}{\fvarCtx}{e}
@@
 $\fbox{$\tKindX{\theta}{\Theta}$}$
-\begin{mathpar}
-\inferrule[op]{
+~~~$\Theta ::= \kOpS{op}{\kappaidx} \pipe \Theta, \Theta$~~~
+\begin{mathpar}
+\inferrule[op-kinding]{
 	\tKind{\tvarCtx, \tOfKind{\tvar{i}}{\kappai}, \tOfKind{\tvar{a}}{\klist{\kDen}}}{\fvarCtx}{\tau}{\kDen}
 }{
-	\opType{\tvarCtx}{\fvarCtx}{{\tops{op}{\kappai}{i}{a}{\tau}}}{
-	\kOpS{op}{\kappai}}
-}
-
-\inferrule[ops]{
+	\opType{\tvarCtx}{\fvarCtx}{{\tops{op}{\kappaidx}{i}{a}{\tau}}}{
+	\kOpS{op}{\kappaidx}}
+}
+
+\inferrule[ops-kinding]{
 	\tKindX{\theta_1}{\Theta_1}\\
@@
 \begin{mathpar}
-\inferrule[e-var-kind]{ }{
+\inferrule[e-var-kinding]{ }{
 	\exprOK{\tvarCtx}{\itvarCtx, \evar{x}}{\fvarCtx}{\evar{x}}
@@
 
-\inferrule[e-lam-kind]{
+\inferrule[e-lam-kinding]{
 	\tKindX{\tau}{\kTypeBlur}\\
@@
 
-\inferrule[e-op-kind]{
+\inferrule[e-op-kinding]{
 	\fvarOfType{Fam}{\kappaidx}{\Theta} \in \fvarCtx\\
@@
 \end{mathpar}
-\caption{\small Kinding for programs. Variable contexts $\tvarCtx$ and $\itvarCtx$ obey standard structural properties. Kinding for type-level terms in Figure \ref{kindtl}.}
+\caption{\small Kinding for programs. Variable contexts $\tvarCtx$ and $\itvarCtx$ obey standard structural properties. Kinding rules for type-level terms are given in Figure \ref{tlkind}.}
 \label{kindprog}
 \end{figure}
+\subsection{Active Typechecking and Translation}
+The second phase of compilation involves typechecking and translating the program to produce an internal term, invoking the user-defined logic encoded in the introduced operator families when needed.
+The rules for active typechecking and translation are given in Fig.
+\ref{att}.
+The context $\fvalCtx$ tracks the operator definitions and representation schemas associated with families in scope.
+Because the logic inside operators must be invoked during this phase, we provide an evaluation semantics for type-level terms in Fig.
+\ref{tleval}.
+
+The key rule is \textsc{att-exp}, which assigns a type, $\tau$, and an \emph{abstract translation}, $\gabs$, to an expression, $e$, as specified by the judgement $\ecompilesX{e}{\tau}{\gabs}$, then \emph{deabstracts} the translation to produce the final translation.
+Deabstraction is specified in Fig.
+\ref{deabs} by the judgement $\eraseX{\gabs}{\gamma}$.
+The purpose of the abstract translation is to ensure that the implementation details of a type cannot be observed by operators associated with a different type, ensuring that invariants maintained by that type cannot be violated by operators in a different type family.
+For example, the implementation of natural numbers as integers in Fig.
+\ref{example} maintains the invariant that the translation is non-negative.
+If the knowledge that natural numbers are implemented as integers was externally visible, a different extension could introduce an operator like $\tops{badnum}{\kunit}{i}{a}{\tapp{\tvar{const}}{\tapp{\tvar{a}}{\tden{\titerm{-1}}{\ttype{Nat}{\tunit}}}}}$.
+
+This is prevented by a mechanism quite similar to the mechanism by which abstract types in ML-style modules operate \cite{atapl} -- by keeping the representation schema hidden outside the type family it is associated with.
+As a result, it cannot be shown given the knowledge available in the type family containing $\opvar{badnum}$ that the internal type associated with $\ttype{Nat}{\tunit}$ is $\dint$, and so $-1$ is not valid according to the rules we will describe below.
+The only way to produce a term of this type in an operator not associated with the type is to extract it from an input argument, where the invariants are guaranteed inductively to have been  maintained.
+We will make this more concrete in Sec.
+\ref{safety}.
+
+Our representation schema abstraction mechanism is specified syntactically, so it relates to previous work on syntactic type abstraction \cite{syntypeabs}.
+While that work was focused on abstracting away the identity of a particular type outside of a ``host'', we focus on abstracting away the knowledge of how a primitive type family is implemented outside of the operators associated with it.
+
+Variables translate directly to variables in the internal language, with the type determined by the typing context, $\Gamma$ (rule \textsc{att-var}).
+Lambda terms translate to lambda terms in the internal language.
+
+\subsection{Abstract Representations}
+Note that there is no elimination form that could be used to determine the internal type associated with a type, but the special internal type $\trepof{\tau}$, where $\tau$ is a type, can be used to refer to it abstractly
+
 \begin{figure}[t]
@@
 \inferrule[att-exp]{
-	\ecompiles{\emptyctx}{\fvalCtx}{e}{\tau}{\ghat}\\
-	\eraseX{\ghat}{\gamma}
+	\ecompiles{\emptyctx}{\fvalCtx}{e}{\tau}{\gabs}\\
+	\eraseX{\gabs}{\gamma}
 }{
@@
 \end{mathpar}
-$\fbox{\inferrule{}{\ecompilesX{e}{\tau}{\ghat}}}$
+$\fbox{\inferrule{}{\ecompilesX{e}{\tau}{\gabs}}}$
 ~~~$\etCtx ::= \emptyctx \pipe \etCtxX{x}{\tau}$
@@
 \inferrule[att-lam]{
-	\tEvalX{\tau_1}{\ttype{Fam}{\tauidx}}\\
-	\ecompiles{\etCtxX{x}{\ttype{Fam}{\tauidx}}}{\fvalCtx}{e}{\tau_2}{\ghat}\\\\
-	\tauisdel{Arrow}{\fvalCtx}{Fam}{\tauidx}{\dbar}\\
-}{
-	\ecompiles{\etCtx}{\fvalCtx}{\elam{\evar{x}}{\tau_1}{e}}{\ttype{Arrow}{\tpair{\ttype{Fam}{\tauidx}}{\tau_2}}}{\ilam{\evar{x}}{\dbar}{\ghat}}
+	\tEvalX{\tau_1}{\tau_1'}\\
+	\delfromtau{$\Xi_0$}{\fvalCtx}{\tau_1'}{\sabs}\\\\
+	\ecompiles{\etCtxX{x}{\tau_1'}}{\fvalCtx}{e}{\tau_2}{\gabs}
+}{
+	\ecompiles{\etCtx}{\fvalCtx}{\elam{\evar{x}}{\tau_1}{e}}{\ttype{Arrow}{\tpair{\tau_1'}{\tau_2}}}{\ilam{\evar{x}}{\sabs}{\gabs}}
 }
@@
 	\tEvalX{\taui}{\taui'}\\\\
-	\ecompiles{\etCtx}{\fvalCtx}{e_1}{\tau_1}{\ghat_1}~~~~
+	\ecompiles{\etCtx}{\fvalCtx}{e_1}{\tau_1}{\gamma_1}~~~~
 	\cdots~~~~
-	\ecompiles{\etCtx}{\fvalCtx}{e_n}{\tau_n}{\ghat_n}\\
+	\ecompiles{\etCtx}{\fvalCtx}{e_n}{\tau_n}{\gamma_n}\\
 	\text{$\begin{array}{c}
@@
 			\taui'/\tvar{i}\\
-			\tden{\titerm{\ghat_1}}{\tau_1} :: \cdots :: \tden{\titerm{\ghat_n}}{\tau_n} :: \tnil{\kDen}/\tvar{a}
+			\tden{\titerm{\gamma_1}}{\tau_1} :: \cdots :: \tden{\titerm{\gamma_n}}{\tau_n} :: \tnil{\kDen}/\tvar{a}
 			\end{array}$}\right]\tauop
 		\Downarrow
-		\tden{\titerm{\ghat}}{\ttype{Fam'}{\tauidx}}
+		\tden{\titerm{\gamma}}{\ttype{Fam'}{\tauidx}}
 		\end{array}$}\\
-	\tauisdel{Fam}{\fvalCtx}{Fam'}{\tauidx}{\dbar}\\
+	\tauisdel{Fam}{\fvalCtx}{Fam'}{\tauidx}{\sigma}\\
 	\eCtxTogCtx{Fam}{\fvalCtx}{\etCtx}{\gtCtx}\\
-	\checkRC{\gtCtx}{Fam}{\fvalCtx}{\ghat}{\dbar}
-}{
-	\ecompiles{\etCtx}{\fvalCtx}{\eop{Fam}{op}{\taui}{\splat{e}{1}{n}}}{\ttype{Fam'}{\tauidx}}{\ghat}
-}
-\end{mathpar}
-\caption{\small Active typechecking and translation}
+	\checkRC{\gtCtx}{Fam}{\fvalCtx}{\gamma}{\sigma}
+}{
+	\ecompiles{\etCtx}{\fvalCtx}{\eop{Fam}{op}{\taui}{\splat{e}{1}{n}}}{\ttype{Fam'}{\tauidx}}{\gamma}
+}
+\end{mathpar}
+\caption{\small Active typechecking and translation of programs. Evaluation semantics for type-level terms are given in Fig. \ref{tleval}}
+\label{att}
 \end{figure}
@@
 \small
-\inferrule[var kind]{
+\inferrule[var-kind]{
 }{
@@
 
-\inferrule[k-arrow intro]{
+\inferrule[k-arrow-intro]{
   \tKind{\tvarCtxX{t}{\kappa_1}}{\fvarCtx}{\tau}{\kappa_2}
@@
 
-\inferrule[k-arrow elim]{
+\inferrule[k-arrow-elim]{
   \tKindX{\tau_1}{\karrow{\kappa_1}{\kappa_2}}\\
@@
 \\
-\text{\color{gray} (standard statics for integers, strings, products and lists omitted)}
+\text{\color{gray} (standard statics for lists, integers, strings and products  omitted)}
 \\
@@
 %% TODO: list
-\inferrule[eq kind]{
+\inferrule[tl-equality]{
 	\kEq{\kappa}\\
@@
 
-\inferrule[type intro]{
+\inferrule[type-intro]{
 	\fvarOfTypeDf \in \fvarCtx\\\\
@@
 
-\inferrule[type elim]{
+\inferrule[type-elim]{
 	\fvarOfTypeDf \in \fvarCtx\\
@@
 }
-
-\inferrule[den intro valid]{
-	\tKindX{\tautype}{\kTypeBlur}\\\\
-	\tKindX{\tauval}{\kITerm}
-}{
-	\tKindX{\tden{\tauval}{\tautype}}{\kDen}
-}
-
-\inferrule[den intro err]{ }{
+~~~~~~
+\inferrule[den-intro-valid]{
+	\tKindX{\tau_1}{\kITerm}\\\\
+	\tKindX{\tau_2}{\kTypeBlur}
+}{
+	\tKindX{\tden{\tau_1}{\tau_2}}{\kDen}
+}
+~~~~~~
+\inferrule[den-intro-err]{ }{
 	\tKindX{\terr}{\kDen}
@@
 
-\inferrule[den elim]{
-	\tKindX{\tauden}{\kDen}\\
-	\tKind{\tvarCtx, \tOfKind{\tvar{x}}{\kITerm}, \tOfKind{\tvar{y}}{\kTypeBlur}}{\fvarCtx}{\tau_1}{\kappa}\\
+\inferrule[den-elim]{
+	\tKindX{\tau}{\kDen}\\\\
+	\tKind{\tvarCtx, \tOfKind{\tvar{x}}{\kITerm}, \tOfKind{\tvar{t}}{\kTypeBlur}}{\fvarCtx}{\tau_1}{\kappa}\\
 	\tKindX{\tau_2}{\kappa}
 }{
-	\tKindX{\tdencase{\tauden}{y}{x}{\tau_1}{\tau_2}}{\kappa}
-}
-
-\inferrule[iterm intro]{
-	\isIterm{\tvarCtx}{\emptyctx}{\fvarCtx}{\ghat}
-}{
-	\tKindX{\titerm{\ghat}}{\kITerm}
+	\tKindX{\tdencase{\tau}{x}{t}{\tau_1}{\tau_2}}{\kappa}
+}
+
+\inferrule[iterm-intro]{
+	\isIterm{\tvarCtx}{\emptyctx}{\fvarCtx}{\gamma}
+}{
+	\tKindX{\titerm{\gamma}}{\kITerm}
 }
@@
 \inferrule[itype intro]{
-	\isItype{\tvarCtx}{\fvarCtx}{\dhat}
-}{
-	\tKindX{\titype{\dhat}}{\kIType}
+	\isItype{\tvarCtx}{\fvarCtx}{\sigma}
+}{
+	\tKindX{\titype{\sigma}}{\kIType}
 }
@@
 
-\inferrule[str-eq]{ }{
+\inferrule[s-eq]{ }{
 	\kEq{\kstr}
@@
 
-\inferrule[prod-eq]{
+\inferrule[p-eq]{
 	\kEq{\kappa_1}\\
@@
 
-\inferrule[list-eq]{
+\inferrule[l-eq]{
 	\kEq{\kappa}
@@
 \end{mathpar}
-$\fbox{$\isItermX{\ghat}$}$
-\begin{mathpar}
-\inferrule[i-var kind]{ }{
+$\fbox{$\isItermX{\gamma}$}$
+\begin{mathpar}
+\inferrule[i-var-kinding]{ }{
 	\isIterm{\tvarCtx}{\eivarCtxX{x}}{\fvarCtx}{\evar{x}}
@@
 
-\inferrule[i-lam kind]{
-	\isItypeX{\dhat}\\\\
-	\isIterm{\tvarCtx}{\eivarCtxX{x}}{\fvarCtx}{\ghat}
-}{
-	\isItermX{\ilam{\evar{x}}{\dhat}{\ghat}}
-}
-
-\inferrule[i-fix kind]{
-	\isItypeX{\dhat}\\\\
-	\isIterm{\tvarCtx}{\eivarCtxX{x}}{\fvarCtx}{\ghat}
-}{
-	\isItermX{\ifix{\evar{f}}{\dhat}{\ghat}}
+\inferrule[i-lam-kinding]{
+	\isItypeX{\sigma}\\\\
+	\isIterm{\tvarCtx}{\eivarCtxX{x}}{\fvarCtx}{\gamma}
+}{
+	\isItermX{\ilam{\evar{x}}{\sigma}{\gamma}}
+}
+
+\inferrule[i-fix-kinding]{
+	\isItypeX{\sigma}\\\\
+	\isIterm{\tvarCtx}{\eivarCtxX{x}}{\fvarCtx}{\gamma}
+}{
+	\isItermX{\ifix{\evar{f}}{\sigma}{\gamma}}
 }
@@
 \\
-\inferrule[iterm unquote kind]{
+\inferrule[iterm-dereify-kinding]{
 	\tKindX{\tau}{\kITerm}
@@
 
-\inferrule[val from den kind]{
-	\tKindX{\tautype}{\kTypeBlur}\\
-	\tKindX{\tauval}{\kITerm}
-}{
-	\isItermX{\tvalof{\tauval}{\tautype}}
-}
-\end{mathpar}
-$\fbox{$\isItypeX{\dhat}$}$
-\begin{mathpar}
-\inferrule[i-int kind]{ }{
+\inferrule[abs-trans-kinding]{
+	\tKindX{\tauiterm}{\kITerm}\\
+	\tKindX{\tautype}{\kTypeBlur}
+}{
+	\isItermX{\tvalof{\tauiterm}{\tautype}}
+}
+\end{mathpar}
+$\fbox{$\isItypeX{\sigma}$}$
+\begin{mathpar}
+\inferrule[i-int-kinding]{ }{
 	\isItypeX{\dint}
@@
 
-\inferrule[i-prod kind]{
-	\isItypeX{\dhat_1}\\\\
-	\isItypeX{\dhat_2}
-}{
-	\isItypeX{\dpair{\dhat_1}{\dhat_2}}
-}
-
-\inferrule[i-arrow kind]{
-	\isItypeX{\dhat_1}\\\\
-	\isItypeX{\dhat_2}
-}{
-	\isItypeX{\darrow{\dhat_1}{\dhat_2}}
-}
-
-\inferrule[itype unquote kind]{
+\inferrule[i-prod-kinding]{
+	\isItypeX{\sigma_1}\\\\
+	\isItypeX{\sigma_2}
+}{
+	\isItypeX{\dpair{\sigma_1}{\sigma_2}}
+}
+
+\inferrule[i-arrow-kinding]{
+	\isItypeX{\sigma_1}\\\\
+	\isItypeX{\sigma_2}
+}{
+	\isItypeX{\darrow{\sigma_1}{\sigma_2}}
+}
+
+\inferrule[itype-dereify-kinding]{
 	\tKindX{\tau}{\kIType}
@@
 
-\inferrule[rep from type kind]{
+\inferrule[abs-rep-kinding]{
 	\tKindX{\tau}{\kTypeBlur}
@@
 \begin{mathpar}
-\inferrule[tl-lam eval]{ }{
+\inferrule[tl-lam-eval]{ }{
 	\tEvalX{\tlam{t}{\kappa}{\tau}}{\tlam{t}{\kappa}{\tau}}
@@
 
-\inferrule[tl-ap eval]{
+\inferrule[tl-ap-eval]{
 	\tEvalX{\tau_1}{\tlam{t}{\kappa}{\tau}}\\
@@
 
-\inferrule[tl-eq eval eq]{
+\inferrule[tl-eq-eval-equal]{
 	\tEvalX{\tau_1}{\tau_1'}\\
@@
 
-\inferrule[tl-eq eval neq]{
+\inferrule[tl-eq-eval-inequal]{
 	\tEvalX{\tau_1}{\tau_1'}\\
@@
 
-\inferrule[type eval]{
+\inferrule[type-eval]{
 	\tEvalX{\tauidx}{\tauidx'}
@@
 
-\inferrule[famcase eval match]{
+\inferrule[famcase-eval-match]{
 	\tEvalX{\tau}{\ttype{Fam}{\tauidx}}\\
@@
 
-\inferrule[famcase eval fail]{
+\inferrule[famcase-eval-fail]{
 	\tEvalX{\tau}{\ttype{Fam'}{\tauidx}}\\
@@
 
-\inferrule[den eval]{
-	\tEvalX{\tautype}{\tautype'}\\
-	\tEvalX{\tauval}{\tauval'}
-}{
-	\tEvalX{\tden{\tauval}{\tautype}}{\tden{\tauval'}{\tautype'}}
-}
-
-\inferrule[err eval]{ }{
+\inferrule[den-valid-eval]{
+	\tEvalX{\tau_1}{\tau_1'}\\
+	\tEvalX{\tau_2}{\tau_2'}
+}{
+	\tEvalX{\tden{\tau_1}{\tau_2}}{\tden{\tau_1'}{\tau_2'}}
+}
+
+\inferrule[den-err-eval]{ }{
 	\tEvalX{\terr}{\terr}
@@
 \inferrule[dencase eval valid]{
-	\tEvalX{\tauden}{\tden{\tauval}{\tautype}}\\
-	\tEvalX{[\tautype/\tvar{x}, \titerm{\tvalof{\tauval}{\tautype}}/\tvar{y}]\tau_1}{\tau_1'}
-}{
-	\tEvalX{\tdencase{\tauden}{y}{x}{\tau_1}{\tau_2}}{\tau_1'}
-}
-
-\inferrule[dencase eval err]{
-	\tEvalX{\tauden}{\terr}\\
+	\tEvalX{\tau}{\tden{\tauiterm}{\tautype}}\\\\
+	\tEvalX{[\titerm{\tvalof{\tauiterm}{\tautype}}/\tvar{x}, \tautype/\tvar{t}]\tau_1}{\tau_1'}
+}{
+	\tEvalX{\tdencase{\tau}{x}{t}{\tau_1}{\tau_2}}{\tau_1'}
+}
+
+\inferrule[dencase-eval-err]{
+	\tEvalX{\tau}{\terr}\\
 	\tEvalX{\tau_2}{\tau_2'}
@@
 
-\inferrule[iterm quote eval]{
-	\tiEvalX{\ghat}{\ghat'}
-}{
-	\tEvalX{\titerm{\ghat}}{\titerm{\ghat'}}
-}
-
-\inferrule[itype quote eval]{
-	\tiEvalX{\dhat}{\dhat'}
-}{
-	\tEvalX{\titype{\dhat}}{\titype{\dhat'}}
-}
-\end{mathpar}
-$\fbox{$\tiEvalX{\ghat}{\ghat'}$}$
-\begin{mathpar}
-\inferrule[i-var eval]{ }{
+\inferrule[iterm-reify]{
+	\tiEvalX{\gamma}{\gamma'}
+}{
+	\tEvalX{\titerm{\gamma}}{\titerm{\gamma'}}
+}
+
+\inferrule[itype-reify]{
+	\tiEvalX{\sigma}{\sigma'}
+}{
+	\tEvalX{\titype{\sigma}}{\titype{\sigma'}}
+}
+\end{mathpar}
+$\fbox{$\tiEvalX{\gamma}{\gamma'}$}$
+\begin{mathpar}
+\inferrule[i-var-eval]{ }{
 	\tiEvalX{\evar{x}}{\evar{x}}
@@
 
-\inferrule[i-lam eval]{
-	\tiEvalX{\dhat}{\dhat'}\\
-	\tiEvalX{\ghat}{\ghat'}
-}{
-	\tiEvalX{\ilam{\evar{x}}{\dhat}{\ghat}}{\ilam{\evar{x}}{\dhat'}{\ghat'}}
-}
-
-\inferrule[i-fix eval]{
-	\tiEvalX{\dhat}{\dhat'}\\
-	\tiEvalX{\ghat}{\ghat'}
-}{
-	\tiEvalX{\ifix{\evar{f}}{\dhat}{\ghat}}{\ifix{\evar{f}}{\dhat'}{\ghat'}}
+\inferrule[i-lam-eval]{
+	\tiEvalX{\sigma}{\sigma'}\\
+	\tiEvalX{\gamma}{\gamma'}
+}{
+	\tiEvalX{\ilam{\evar{x}}{\sigma}{\gamma}}{\ilam{\evar{x}}{\sigma'}{\gamma'}}
+}
+
+\inferrule[i-fix-eval]{
+	\tiEvalX{\sigma}{\sigma'}\\
+	\tiEvalX{\gamma}{\gamma'}
+}{
+	\tiEvalX{\ifix{\evar{f}}{\sigma}{\gamma}}{\ifix{\evar{f}}{\sigma'}{\gamma'}}
 }
@@
 \inferrule[val from den eval]{
-	\tEvalX{\tautype}{\tautype'}\\
-	\tEvalX{\tauval}{\tauval'}
-}{
-	\tiEvalX{\tvalof{\tauval}{\tautype}}{\tvalof{\tauval'}{\tautype'}}
+	\tEvalX{\tauiterm}{\tauiterm'}\\
+	\tEvalX{\tautype}{\tautype'}
+}{
+	\tiEvalX{\tvalof{\tauiterm}{\tautype}}{\tvalof{\tauiterm'}{\tautype'}}
 }
@@
 \begin{mathpar}
-\inferrule[i-int eval]{ }{
+\inferrule[i-int-eval]{ }{
 	\tiEvalX{\dint}{\dint}
@@
 
-\inferrule[i-prod eval]{
-	\tiEvalX{\dhat_1}{\dhat_1'}\\
-	\tiEvalX{\dhat_2}{\dhat_2'}
-}{
-	\tiEvalX{\dpair{\dhat_1}{\dhat_2}}{\dpair{\dhat_1'}{\dhat_2'}}
-}
-
-\inferrule[i-arrow eval]{
-	\tiEvalX{\dhat_1}{\dhat_1'}\\
-	\tiEvalX{\dhat_2}{\dhat_2'}
-}{
-	\tiEvalX{\darrow{\dhat_1}{\dhat_2}}{\darrow{\dhat_1'}{\dhat_2'}}
-}
-
-\inferrule[itype unquote eval]{
+\inferrule[i-prod-eval]{
+	\tiEvalX{\sigma_1}{\sigma_1'}\\
+	\tiEvalX{\sigma_2}{\sigma_2'}
+}{
+	\tiEvalX{\dpair{\sigma_1}{\sigma_2}}{\dpair{\sigma_1'}{\sigma_2'}}
+}
+
+\inferrule[i-arrow-eval]{
+	\tiEvalX{\sigma_1}{\sigma_1'}\\
+	\tiEvalX{\sigma_2}{\sigma_2'}
+}{
+	\tiEvalX{\darrow{\sigma_1}{\sigma_2}}{\darrow{\sigma_1'}{\sigma_2'}}
+}
+\\
+\inferrule[itype-eval]{
 	\tEvalX{\tau}{\tau'}
@@
 
-\inferrule[rep from type eval]{
+\inferrule[abs-rep-eval]{
 	\tEvalX{\tau}{\tau'}
@@
 \small
-$\fbox{\inferrule{}{\delfromtau{Fam}{\fvalCtx}{\tau}{\dbar}}}$
+$\fbox{\inferrule{}{\ddbarX{\sigma}{\sigma'}}}$
 ~~~$\fvalCtx ::= \fvalCtx_0 \pipe \fvalCtx, \fvalDf$
+\begin{mathpar}
+\inferrule[abs int]{ }{
+	\ddbarX{\dint}{\dint}
+}
+
+\inferrule[abs arrow]{
+	\ddbarX{\sigma_1}{\sigma_1'}\\
+	\ddbarX{\sigma_2}{\sigma_2'}
+}{
+	\ddbarX{\darrow{\sigma_1}{\sigma_2}}{\darrow{\sigma_1'}{\sigma_2'}}
+}
+
+\inferrule[abs prod]{
+	\ddbarX{\sigma_1}{\sigma_1'}\\
+	\ddbarX{\sigma_2}{\sigma_2'}
+}{
+	\ddbarX{\dpair{\sigma_1}{\sigma_2}}{\dpair{\sigma_1'}{\sigma_2'}}
+}
+
+\inferrule[abs+cancel unquote]{
+	\ddbarX{\sigma}{\sigma'}
+}{
+	\ddbarX{\dup{\titype{\sigma}}}{\sigma'}
+}
+
+\inferrule[abs rep from type visible]{
+	\fvalDf \in \fvalCtx\\
+	\tEvalX{[\tauidx/\tvar{i}]\tau}{\titype{\sigma}}\\
+	\ddbarX{\sigma}{\sigma}
+}{
+	\ddbarX{\trepof{\ttype{Fam}{\tauidx}}}{\sigma}
+}
+
+\inferrule[abs rep from type hidden]{
+	\fvar{Fam} \neq \fvar{Fam'}
+}{
+	\ddbarX{\trepof{\ttype{Fam'}{\tauidx}}}{\trepof{\ttype{Fam'}{\tauidx}}}
+}
+\end{mathpar}
+$\fbox{\inferrule{}{\delfromtau{Fam}{\fvalCtx}{\tau}{\sigma}}}$
 \begin{mathpar}
@@
 	\fval{Fam'}{\theta}{i}{\tau} \in \fvalCtx\\
-	\tEvalX{[\tauidx/\tvar{i}]\tau}{\titype{\dhat}}\\
-	\ddbarX{\dhat}{\dbar}
-}{
-	\tauisdel{Fam}{\fvalCtx}{Fam'}{\tauidx}{\dbar}
-}
-\end{mathpar}
-$\fbox{\inferrule{}{\ddbarX{\dhat}{\dbar}}}$
-\begin{mathpar}
-\inferrule[abs int]{ }{
-	\ddbarX{\dint}{\dint}
-}
-
-\inferrule[abs arrow]{
-	\ddbarX{\dhat_1}{\dbar_1}\\
-	\ddbarX{\dhat_2}{\dbar_2}
-}{
-	\ddbarX{\darrow{\dhat_1}{\dhat_2}}{\darrow{\dbar_1}{\dbar_2}}
-}
-
-\inferrule[abs prod]{
-	\ddbarX{\dhat_1}{\dbar_1}\\
-	\ddbarX{\dhat_2}{\dbar_2}
-}{
-	\ddbarX{\dpair{\dhat_1}{\dhat_2}}{\dpair{\dbar_1}{\dbar_2}}
-}
-
-\inferrule[abs+cancel unquote]{
-	\ddbarX{\dhat}{\dbar}
-}{
-	\ddbarX{\dup{\titype{\dhat}}}{\dbar}
-}
-
-\inferrule[abs rep from type visible]{
-	\fvalDf \in \fvalCtx\\
-	\tEvalX{[\tauidx/\tvar{i}]\tau}{\titype{\dhat}}\\
-	\ddbarX{\dhat}{\dbar}
-}{
-	\ddbarX{\trepof{\ttype{Fam}{\tauidx}}}{\dbar}
-}
-
-\inferrule[abs rep from type hidden]{
-	\fvar{Fam} \neq \fvar{Fam'}
-}{
-	\ddbarX{\trepof{\ttype{Fam'}{\tauidx}}}{\trepof{\ttype{Fam'}{\tauidx}}}
+	\tEvalX{[\tauidx/\tvar{i}]\tau}{\titype{\sigma}}\\
+	\ddbarX{\sigma}{\sigma}
+}{
+	\tauisdel{Fam}{\fvalCtx}{Fam'}{\tauidx}{\sigma}
 }
@@
 $\fbox{\inferrule{}{\eCtxTogCtxX{\etCtx}{\gtCtx}}}$
-~~~$\gtCtx ::= \emptyctx \pipe \gtCtxX{x}{\dbar}$
+~~~$\gtCtx ::= \emptyctx \pipe \gtCtxX{x}{\sigma}$
 \begin{mathpar}
@@
 	\eCtxTogCtxX{\etCtx}{\gtCtx}\\
-	\delfromtau{Fam}{\fvalCtx}{\tau}{\dbar}
-}{
-	\eCtxTogCtxX{\etCtxX{x}{\tau}}{\gtCtxX{x}{\dbar}}
-}
-\end{mathpar}
-$\fbox{\inferrule{}{\checkRCX{\ghat}{\dbar}}}$
+	\delfromtau{Fam}{\fvalCtx}{\tau}{\sigma}
+}{
+	\eCtxTogCtxX{\etCtxX{x}{\tau}}{\gtCtxX{x}{\sigma}}
+}
+\end{mathpar}
+$\fbox{\inferrule{}{\checkRCX{\gamma}{\sigma}}}$
 \begin{mathpar}
 \inferrule[abs i-var]{ }{
-	\checkRC{\gtCtxX{x}{\dbar}}{Fam}{\fvalCtx}{\evar{x}}{\dbar}
+	\checkRC{\gtCtxX{x}{\sigma}}{Fam}{\fvalCtx}{\evar{x}}{\sigma}
 }
@@
 \inferrule[abs i-lam]{
-	\ddbarX{\dhat_1}{\dbar_1}\\
-	\checkRC{\gtCtxX{x}{\dbar_1}}{Fam}{\fvalCtx}{\gamma}{\dbar_2}
-}{
-	\checkRCX{\ilam{x}{\dhat_1}{\gamma}}{\darrow{\dbar_1}{\dbar_2}}
+	\ddbarX{\sigma_1}{\sigma_1}\\
+	\checkRC{\gtCtxX{x}{\sigma_1}}{Fam}{\fvalCtx}{\gamma}{\sigma_2}
+}{
+	\checkRCX{\ilam{x}{\sigma_1}{\gamma}}{\darrow{\sigma_1}{\sigma_2}}
 }
@@
 \inferrule[abs i-ap]{
-	\checkRCX{\ghat_1}{\darrow{\dbar_1}{\dbar_2}}\\\\
-	\checkRCX{\ghat_2}{\dbar_1}
-}{
-	\checkRCX{\iapp{\ghat_1}{\ghat_2}}{\dbar_2}
+	\checkRCX{\gamma_1}{\darrow{\sigma_1}{\sigma_2}}\\
+	\checkRCX{\gamma_2}{\sigma_1}
+}{
+	\checkRCX{\iapp{\gamma_1}{\gamma_2}}{\sigma_2}
 }
@@
 \inferrule[abs i-fix]{
-	\ddbarX{\dhat}{\dbar}\\
-	\checkRC{\gtCtxX{x}{\dbar}}{Fam}{\fvalCtx}{\ghat}{\dbar}
-}{
-	\checkRCX{\ifix{x}{\dhat}{\ghat}}{\dbar}
+	\ddbarX{\sigma}{\sigma}\\
+	\checkRC{\gtCtxX{x}{\sigma}}{Fam}{\fvalCtx}{\gamma}{\sigma}
+}{
+	\checkRCX{\ifix{x}{\sigma}{\gamma}}{\sigma}
 }
@@
 %\inferrule[abs pair]{
-%	\checkRCX{\gamma_1}{\dbar_1}\\
-%	\checkRCX{\gamma_2}{\dbar_2}
+%	\checkRCX{\gamma_1}{\sigma_1}\\
+%	\checkRCX{\gamma_2}{\sigma_2}
 %}{
-%	\checkRCX{\ipair{\gamma_1}{\gamma_2}}{\dpair{\dbar_1}{\dbar_2}}
+%	\checkRCX{\ipair{\gamma_1}{\gamma_2}}{\dpair{\sigma_1}{\sigma_2}}
 %}
@@
 %\inferrule[abs fst]{
-%	\checkRCX{\gamma}{\dpair{\dbar_1}{\dbar_2}}
+%	\checkRCX{\gamma}{\dpair{\sigma_1}{\sigma_2}}
 %}{
-%	\checkRCX{\ifst{\gamma}}{\dbar_1}
+%	\checkRCX{\ifst{\gamma}}{\sigma_1}
 %}
@@
 %\inferrule[abs snd]{
-%	\checkRCX{\gamma}{\dpair{\dbar_1}{\dbar_2}}
+%	\checkRCX{\gamma}{\dpair{\sigma_1}{\sigma_2}}
 %}{
-%	\checkRCX{\isnd{\gamma}}{\dbar_2}
+%	\checkRCX{\isnd{\gamma}}{\sigma_2}
 %}
@@
 \inferrule[abs if eq]{
-	\checkRCX{\ghat_1}{\dint}\\
-	\checkRCX{\ghat_2}{\dint}\\\\
-	\checkRCX{\ghat_3}{\dbar}\\
-	\checkRCX{\ghat_4}{\dbar}
-}{
-	\checkRCX{\iIfEq{\ghat_1}{\ghat_2}{\dint}{\ghat_3}{\ghat_4}}{\dbar}
+	\checkRCX{\gamma_1}{\dint}\\
+	\checkRCX{\gamma_2}{\dint}\\\\
+	\checkRCX{\gamma_3}{\sigma}\\
+	\checkRCX{\gamma_4}{\sigma}
+}{
+	\checkRCX{\iIfEq{\gamma_1}{\gamma_2}{\dint}{\gamma_3}{\gamma_4}}{\sigma}
 }
@@
 \inferrule[abs iterm unquote]{
-	\checkRCX{\ghat}{\dbar}
-}{
-	\checkRCX{\iup{\titerm{\ghat}}}{\dbar}
+	\checkRCX{\gamma}{\sigma}
+}{
+	\checkRCX{\iup{\titerm{\gamma}}}{\sigma}
 }
@@
 \inferrule[abs val from den visible]{
-	\tauisdel{Fam}{\fvalCtx}{Fam}{\tauidx}{\dbar}\\
-	\checkRCX{\ghat}{\dbar}
-}{
-	\checkRCX{\tvalof{\titerm{\ghat}}{\ttype{Fam}{\tauidx}}}{\dbar}
+	\tauisdel{Fam}{\fvalCtx}{Fam}{\tauidx}{\sigma}\\
+	\checkRCX{\gamma}{\sigma}
+}{
+	\checkRCX{\tvalof{\titerm{\gamma}}{\ttype{Fam}{\tauidx}}}{\sigma}
 }
@@
 	\fvar{Fam} \neq \fvar{Fam'}\\
-	\tauisdel{Fam}{\fvalCtx}{Fam'}{\tauidx}{\dbar}\\
-	\checkRCX{\ghat}{\dbar}
-}{
-	\checkRCX{\tvalof{\titerm{\ghat}}{\ttype{Fam'}{\tauidx}}}{\trepof{\ttype{Fam'}{\tauidx}}}
+	\tauisdel{Fam}{\fvalCtx}{Fam'}{\tauidx}{\sigma}\\
+	\checkRCX{\gamma}{\sigma}
+}{
+	\checkRCX{\tvalof{\titerm{\gamma}}{\ttype{Fam'}{\tauidx}}}{\trepof{\ttype{Fam'}{\tauidx}}}
 }
@@
 \small
-$\fbox{\inferrule{}{\eraseX{\dbar}{\sigma}}}$
+$\fbox{\inferrule{}{\eraseX{\sigma}{\sigma}}}$
 \begin{mathpar}
@@
 \inferrule[deabs arrow]{
-	\eraseX{\dbar_1}{\sigma_1}\\\\
-	\eraseX{\dbar_2}{\sigma_2}
-}{
-	\eraseX{\darrow{\dbar_1}{\dbar_2}}{\darrow{\sigma_1}{\sigma_2}}
+	\eraseX{\sigma_1}{\sigma_1}\\\\
+	\eraseX{\sigma_2}{\sigma_2}
+}{
+	\eraseX{\darrow{\sigma_1}{\sigma_2}}{\darrow{\sigma_1}{\sigma_2}}
 }
@@
 \inferrule[deabs prod]{
-	\eraseX{\dbar_1}{\sigma_1}\\\\
-	\eraseX{\dbar_2}{\sigma_2}
-}{
-	\eraseX{\dpair{\dbar_1}{\dbar_2}}{\dpair{\sigma_1}{\sigma_2}}
+	\eraseX{\sigma_1}{\sigma_1}\\\\
+	\eraseX{\sigma_2}{\sigma_2}
+}{
+	\eraseX{\dpair{\sigma_1}{\sigma_2}}{\dpair{\sigma_1}{\sigma_2}}
 }
@@
 	\fvalDf \in \fvalCtx\\
-	\tEvalX{[\tauidx/\tvar{i}]\tau}{\titype{\dhat}}\\\\
-	\ddbarX{\dhat}{\dbar}\\
-	\eraseX{\dbar}{\sigma}
+	\tEvalX{[\tauidx/\tvar{i}]\tau}{\titype{\sigma}}\\\\
+	\ddbarX{\sigma}{\sigma}\\
+	\eraseX{\sigma}{\sigma}
 }{
@@
 \end{mathpar}
-$\fbox{\inferrule{}{\eraseX{\ghat}{\gamma}}}$
+$\fbox{\inferrule{}{\eraseX{\gamma}{\gamma}}}$
 \begin{mathpar}
@@
 \inferrule[deabs i-lam]{
-	\ddbar{\_}{\fvalCtx}{\dhat}{\dbar}\\\\
-	\eraseX{\dbar}{\sigma}\\\\
-	\eraseX{\ghat}{\gamma}
-}{
-	\eraseX{\ilam{x}{\dhat}{\ghat}}{\ilam{x}{\sigma}{\gamma}}
+	\ddbar{\_}{\fvalCtx}{\sigma}{\sigma}\\\\
+	\eraseX{\sigma}{\sigma}\\\\
+	\eraseX{\gamma}{\gamma}
+}{
+	\eraseX{\ilam{x}{\sigma}{\gamma}}{\ilam{x}{\sigma}{\gamma}}
 }
@@
 %\inferrule[deabs ap]{
-%	\eraseX{\ghat_1}{\ghat_1}\\
-%	\eraseX{\ghat_2}{\ghat_2}
+%	\eraseX{\gamma_1}{\gamma_1}\\
+%	\eraseX{\gamma_2}{\gamma_2}
 %}{
-%	\eraseX{\iapp{\ghat_1}{\ghat_2}}{\iapp{\ghat_1}{\ghat_2}}
+%	\eraseX{\iapp{\gamma_1}{\gamma_2}}{\iapp{\gamma_1}{\gamma_2}}
 %}
@@
 \inferrule[deabs i-fix]{
-	\ddbar{\_}{\fvalCtx}{\dhat}{\dbar}\\\\
-	\eraseX{\dbar}{\sigma}\\\\
-	\eraseX{\ghat}{\gamma}
-}{
-	\eraseX{\ifix{x}{\dhat}{\ghat}}{\ifix{x}{\sigma}{\gamma}}
+	\ddbar{\_}{\fvalCtx}{\sigma}{\sigma}\\\\
+	\eraseX{\sigma}{\sigma}\\\\
+	\eraseX{\gamma}{\gamma}
+}{
+	\eraseX{\ifix{x}{\sigma}{\gamma}}{\ifix{x}{\sigma}{\gamma}}
 }
@@
 \inferrule[deabs unquote]{
-	\eraseX{\ghat}{\gamma}
-}{
-	\eraseX{\iup{\titerm{\ghat}}}{\gamma}
+	\eraseX{\gamma}{\gamma}
+}{
+	\eraseX{\iup{\titerm{\gamma}}}{\gamma}
 }
@@
 \inferrule[deabs val from den hidden]{
-	\eraseX{\ghat}{\gamma}
-}{
-	\eraseX{\tvalof{\titerm{\ghat}}{\ttype{Fam}{\tauidx}}}{\gamma}
+	\eraseX{\gamma}{\gamma}
+}{
+	\eraseX{\tvalof{\titerm{\gamma}}{\ttype{Fam}{\tauidx}}}{\gamma}
 }

### 2013-10-12 esop14/att-esop14.tex
Up through abs reps

@@
 \subsection{Active Typechecking and Translation}
+\begin{figure}[t]
+\small
+$\fbox{\inferrule{}{\pcompilesX{\rho}}}$
+~~~$\fvalCtx ::= \fvalCtx_0 \pipe \fvalCtxX{\fvalDf}$
+\begin{mathpar}
+\inferrule[att-fam]{
+	\pcompiles{\fvalCtxX{\fvalDf}}{\rho}{\gamma}
+}{
+	\pcompilesX{\pfam{\familyDf}{\rho}}
+}
+
+\inferrule[att-def]{
+	\tEvalX{\tau}{\tau'}\\
+	\pcompilesX{\subst{\tau'}{\tvar{t}}{\rho}}
+}{
+	\pcompilesX{\pdef{t}{\kappa}{\tau}{\rho}}
+}
+
+\inferrule[att-exp]{
+	\ecompiles{\emptyctx}{\fvalCtx}{e}{\tau}{\gabs}\\
+	\eraseX{\gabs}{\gamma}
+}{
+	\pcompilesX{e}
+}
+\end{mathpar}
+$\fbox{\inferrule{}{\ecompilesX{e}{\tau}{\gabs}}}$
+~~~$\etCtx ::= \emptyctx \pipe \etCtxX{x}{\tau}$
+\begin{mathpar}
+\inferrule[att-var]{ }{
+	\ecompiles{\etCtxX{x}{\tau}}{\fvalCtx}{\evar{x}}{\tau}{\evar{x}}
+}
+
+\inferrule[att-lam]{
+	\tEvalX{\tau_1}{\tau_1'}\\
+	%\delfromtau{$\Xi_0$}{\fvalCtx}{\tau_1'}{\sabs}\\\\
+	\ecompiles{\etCtxX{x}{\tau_1'}}{\fvalCtx}{e}{\tau_2}{\gabs}
+}{
+	\ecompiles{\etCtx}{\fvalCtx}{\elam{\evar{x}}{\tau_1}{e}}{\ttype{Arrow}{\tpair{\tau_1'}{\tau_2}}}{\ilam{\evar{x}}{\trepof{\tau_1'}}{\gabs}}
+}
+
+\inferrule[att-op]{
+	\fvalDf \in \fvalCtx\\
+	\tops{op}{\kappai}{i}{a}{\tauop} \in \theta\\
+	\tEvalX{\tauidx}{\tauidx'}\\\\
+	\ecompiles{\etCtx}{\fvalCtx}{e_1}{\tau_1}{\gamma_1}~~~~
+	\cdots~~~~
+	\ecompiles{\etCtx}{\fvalCtx}{e_n}{\tau_n}{\gamma_n}\\
+	\text{$\begin{array}{c}
+		\left[\text{$\begin{array}{r}
+			\tauidx'/\tvar{i}\\
+			\tden{\titerm{\gamma_1}}{\tau_1} :: \cdots :: \tden{\titerm{\gamma_n}}{\tau_n} :: \tnil{\kDen}/\tvar{a}
+			\end{array}$}\right]\tauop
+		\Downarrow
+		\tden{\titerm{\gabs}}{\tau}
+		\end{array}$}\\
+	\delfromtau{Fam}{\fvalCtx}{\tau}{\sabs}\\
+	\eCtxTogCtx{Fam}{\fvalCtx}{\etCtx}{\gtCtx}\\
+	\checkRC{\gtCtx}{Fam}{\fvalCtx}{\gabs}{\sabs}
+}{
+	\ecompiles{\etCtx}{\fvalCtx}{\eop{Fam}{op}{\tauidx}{\splat{e}{1}{n}}}{\ttype{Fam'}{\tauidx}}{\gabs}
+}
+\end{mathpar}
+\caption{\small Active typechecking and translation of programs. Evaluation semantics for type-level terms are given in Fig. \ref{tleval}}
+\label{att}
+\end{figure}
 The second phase of compilation involves typechecking and translating the program to produce an internal term, invoking the user-defined logic encoded in the introduced operator families when needed.
@@
 \ref{deabs} by the judgement $\eraseX{\gabs}{\gamma}$.
-The purpose of the abstract translation is to ensure that the implementation details of a type cannot be observed by operators associated with a different type, ensuring that invariants maintained by that type cannot be violated by operators in a different type family.
-For example, the implementation of natural numbers as integers in Fig.
+The purpose of the abstract translation is to ensure that the implementation details of a type cannot be observed by operators associated with a different type, which ensures that that invariants maintained by one type family are preserved even if other type families are subsequently introduced.
+For example, our implementation of natural numbers as integers in Fig.
 \ref{example} maintains the invariant that the translation is non-negative.
-If the knowledge that natural numbers are implemented as integers was externally visible, a different extension could introduce an operator like $\tops{badnum}{\kunit}{i}{a}{\tapp{\tvar{const}}{\tapp{\tvar{a}}{\tden{\titerm{-1}}{\ttype{Nat}{\tunit}}}}}$.
+If the knowledge that natural numbers are implemented as integers was externally visible, a different extension could introduce an operator like $\tops{badnum}{\kunit}{i}{a}{\tapp{\tvar{const}}{\tapp{\tvar{a}}{\tden{\titerm{-1}}{\ttype{Nat}{\tunit}}}}}$, breaking this invariant (and thus the guarantee that the recursor always terminates!)
+
+To explain this mechanism, let us begin by going over how expressions are assigned types and abstract translations.
+Variables  (rule \textsc{att-var}) translate directly to variables in the internal language, with the type determined by the typing context, $\Gamma$.
+Lambda terms (rule \textsc{att-lam}) translate to lambda terms in the internal language and are given the appropriate $\fvar{Arrow}$ type.
+The internal type assigned to the variable of the internal lambda term is $\trepof{\tau'}$, which is the abstract internal type associated with $\tau'$, the type of the variable in the lambda expression, $\tau$, after it has been fully evaluated to a type-level value.
+This will be converted to the actual internal type determined by $\tau'$'s representation schema during deabstraction.
 
@@
 As a result, it cannot be shown given the knowledge available in the type family containing $\opvar{badnum}$ that the internal type associated with $\ttype{Nat}{\tunit}$ is $\dint$, and so $-1$ is not valid according to the rules we will describe below.
-The only way to produce a term of this type in an operator not associated with the type is to extract it from an input argument, where the invariants are guaranteed inductively to have been  maintained.
-We will make this more concrete in Sec.
+The only way to produce a term of type $\ttype{Nat}{\tunit}$ in an operator not associated with the family $\fvar{Nat}$ is to extract it from an argument (e.g.
+by projecting it out of an $n$-tuple), so the invariants are guaranteed inductively to have been  maintained.
+We will make this non-interference guarantee more precise in Sec.
 \ref{safety}.
 
-Our representation schema abstraction mechanism is specified syntactically, so it relates to previous work on syntactic type abstraction \cite{syntypeabs}.
-While that work was focused on abstracting away the identity of a particular type outside of a ``host'', we focus on abstracting away the knowledge of how a primitive type family is implemented outside of the operators associated with it.
-
-Variables translate directly to variables in the internal language, with the type determined by the typing context, $\Gamma$ (rule \textsc{att-var}).
-Lambda terms translate to lambda terms in the internal language.
+Our representation schema abstraction mechanism is purely syntactic, so it relates to previous work on syntactic type abstraction \cite{syntypeabs}.
+While that work was focused on abstracting away the identity of a particular type outside of the ``principal'' it is associated with, we focus on abstracting away the knowledge of how a primitive type family is implemented outside of its operators.
 
@@
 
-\begin{figure}[t]
-\small
-$\fbox{\inferrule{}{\pcompilesX{\rho}}}$
-~~~$\fvalCtx ::= \fvalCtx_0 \pipe \fvalCtxX{\fvalDf}$
-\begin{mathpar}
-\inferrule[att-fam]{
-	\pcompiles{\fvalCtxX{\fvalDf}}{\rho}{\gamma}
-}{
-	\pcompilesX{\pfam{\familyDf}{\rho}}
-}
-
-\inferrule[att-def]{
-	\tEvalX{\tau}{\tau'}\\
-	\pcompilesX{\subst{\tau'}{\tvar{t}}{\rho}}
-}{
-	\pcompilesX{\pdef{t}{\kappa}{\tau}{\rho}}
-}
-
-\inferrule[att-exp]{
-	\ecompiles{\emptyctx}{\fvalCtx}{e}{\tau}{\gabs}\\
-	\eraseX{\gabs}{\gamma}
-}{
-	\pcompilesX{e}
-}
-\end{mathpar}
-$\fbox{\inferrule{}{\ecompilesX{e}{\tau}{\gabs}}}$
-~~~$\etCtx ::= \emptyctx \pipe \etCtxX{x}{\tau}$
-\begin{mathpar}
-\inferrule[att-var]{ }{
-	\ecompiles{\etCtxX{x}{\tau}}{\fvalCtx}{\evar{x}}{\tau}{\evar{x}}
-}
-
-\inferrule[att-lam]{
-	\tEvalX{\tau_1}{\tau_1'}\\
-	\delfromtau{$\Xi_0$}{\fvalCtx}{\tau_1'}{\sabs}\\\\
-	\ecompiles{\etCtxX{x}{\tau_1'}}{\fvalCtx}{e}{\tau_2}{\gabs}
-}{
-	\ecompiles{\etCtx}{\fvalCtx}{\elam{\evar{x}}{\tau_1}{e}}{\ttype{Arrow}{\tpair{\tau_1'}{\tau_2}}}{\ilam{\evar{x}}{\sabs}{\gabs}}
-}
-
-\inferrule[att-op]{
-	\fvalDf \in \fvalCtx\\
-	\tops{op}{\kappai}{i}{a}{\tauop} \in \theta\\
-	\tEvalX{\taui}{\taui'}\\\\
-	\ecompiles{\etCtx}{\fvalCtx}{e_1}{\tau_1}{\gamma_1}~~~~
-	\cdots~~~~
-	\ecompiles{\etCtx}{\fvalCtx}{e_n}{\tau_n}{\gamma_n}\\
-	\text{$\begin{array}{c}
-		\left[\text{$\begin{array}{r}
-			\taui'/\tvar{i}\\
-			\tden{\titerm{\gamma_1}}{\tau_1} :: \cdots :: \tden{\titerm{\gamma_n}}{\tau_n} :: \tnil{\kDen}/\tvar{a}
-			\end{array}$}\right]\tauop
-		\Downarrow
-		\tden{\titerm{\gamma}}{\ttype{Fam'}{\tauidx}}
-		\end{array}$}\\
-	\tauisdel{Fam}{\fvalCtx}{Fam'}{\tauidx}{\sigma}\\
-	\eCtxTogCtx{Fam}{\fvalCtx}{\etCtx}{\gtCtx}\\
-	\checkRC{\gtCtx}{Fam}{\fvalCtx}{\gamma}{\sigma}
-}{
-	\ecompiles{\etCtx}{\fvalCtx}{\eop{Fam}{op}{\taui}{\splat{e}{1}{n}}}{\ttype{Fam'}{\tauidx}}{\gamma}
-}
-\end{mathpar}
-\caption{\small Active typechecking and translation of programs. Evaluation semantics for type-level terms are given in Fig. \ref{tleval}}
-\label{att}
-\end{figure}
 \begin{figure}[t]

### 2013-10-12 esop14/att-esop14.tex
in my mind clearer, and a bit shorter, didn't think the bit at the end added a whole lot

--- 2013-10-12 esop14/att-esop14.tex
@@
 For example, regular expressions encoded using inductive datatypes are often considered overly verbose, so most functional languages support them via strings, which is less safe.
-Finally, general-purpose abstractions are implemented in a uniform manner.
-Domain knowledge is not easily applied to eliminate overhead or perform optimizations, and implementations designed for typical application workloads may not be satisfactory in parts of a program where performance is a key criteria, particularly when targeting heterogeneous hardware platforms (e.g.
+Finally, general-purpose abstractions are implemented in a uniform manner, making it more difficult for domain knowledge to be applied to eliminate overhead or perform optimizations, leading to unsatisfactory performance in performance-critical applications -- particularly when targeting heterogeneous hardware platforms (e.g.
+programmable GPUs) and distributed computing resources.
+
+% Domain knowledge is not easily applied to eliminate overhead or perform optimizations, and implementations designed for typical application workloads may not be satisfactory in parts of a program where performance is a key criteria, particularly when targeting heterogeneous hardware platforms (e.g.
 programmable GPUs) and distributed computing resources.

### 2013-10-12 esop14/att-esop14.tex
already kind of discussed in gchat

@@
 This discourages languages from including constructs that rely on statically-checked invariants stronger than those supported by their underlying implementation in the common intermediate language.
-At best, constructs like these can be exposed by generating a wrapper where run-time checks have been inserted to guarantee these invariants.
-This compromises both verifiability and performance.
-For%requires the development of an interoperability layer for every pair of DSLs.
-Moreover, this approach exposes the internals of an implementation to clients, making the abstraction awkward to work with and causing code breakage when implementation details change.
-This defeats a primary purpose of high-level programming languages: hiding low-level details from clients of an abstraction.
+At best, constructs like these can be exposed by generating a wrapper where run-time checks have been inserted to guarantee these invariants, thus compromising both verifiability and performance.
+%requires the development of an interoperability layer for every pair of DSLs. 
+%Moreover, this approach exposes the internals of an implementation to clients, making the abstraction awkward to work with and causing code breakage when implementation details change.
+This defeats a primary purpose of high-level programming languages: hiding low-level details from clients of an abstraction
 We diagram this fundamental \emph{interoperability problem} in Figure \ref{approaches}(a). 
-As an example, F\#'s type system prevents \lstinline{null} values from occurring within data structures, but because it's type system is not available when calling into F\# code from another language, like C\#, run-time null checks must still be included in the implementation.
+%As an example, F\#'s type system prevents \lstinline{null} values from occurring within data structures, but because it's type system is not available when calling into F\# code from another language, like C\#, run-time null checks must still be included in the implementation.
 \begin{figure}[t]

### 2013-10-12 esop14/att-esop14.tex
removed the bit about section 6 because i'm pretty sure it's not gonna exist
and you don't want to forget it in the introduction if it doesn't end
up existing...
i don't think i got much with this one. like 7 characters bleh

@@
 Our choice of a simply-typed, simply-kinded calculus where expressions are given meaning by translation to a simply-typed internal language appears to occupy a ``sweet spot'' in the design space, and relates closely to how simply-typed functional languages like  ML and Haskell are specified and implemented today.
-In Section \ref{design}, we briefly discuss other points in the design space of actively-typed languages and describe the sorts of abstractions that the mechanism as we have introduced it is not capable of expressing, suggesting several directions for future research.
+%In Section \ref{design}, we briefly discuss other points in the design space of actively-typed languages and describe the sorts of abstractions that the mechanism as we have introduced it is not capable of expressing, suggesting several directions for future research.
 We conclude with a discussion of related work in Section \ref{related-work}.
+
 %dependently-typed and object-oriented type-level languages, as well as the constraints governing the design of the internal language.  
@@
 \section{From Extensible Compilers to Extensible Languages}\label{evolution}
-To understand the genesis of our internal extension mechanism, it is helpful to begin by considering why most implementations of programming languages cannot even be  externally extended. 
+To understand the genesis of our internal extension mechanism, we begin by considering why most implementations of programming languages cannot even be  externally extended. 
 Let us consider, as a simple example, an implementation of G\"odel's T, a typed lambda calculus with recursion on primitive natural numbers (see Appendix \ref{T}). 

### 2013-10-12 esop14/att-esop14.tex
i just toned this down, i don't think i really made it much shorter

@@
 If our newly-introduced constructs are exposed at a library's interface boundary, clients using different compilers face the same problems with interoperability that those using different languages face.
-That is, \textbf{extending a language by extending a single compiler for it is morally equivalent to creating a new language}.
-Several prominent language ecosystems today are in a state where a prominent compiler has introduced or enabled the introduction of extensions that many libraries have come to rely on, including the Glasgow Haskell Compiler, SML/NJ and the GNU compilers for C and C++.
-We argue that this practice should be considered harmful.
+That is, extending a language by extending a single compiler for it is morally equivalent to creating a new language.
+Several prominent language ecosystems today are in a state where a prominent compiler has introduced or enabled the introduction of extensions that many libraries have come to rely on, including the Glasgow Haskell Compiler, SML/NJ and the GNU compilers for C and C++, thus complicating interoperability.
 

### 2013-10-12 esop14/att-esop14.tex
all of this talk about control was really unclear to me. i don't think it added much
without it, i think it's a whole lot better. actually, removing the
"without allowing so much control that the reliability of the system
suffers" part made the two sentences connect much better. besides, i
think that that subsentence is a bit hand wavy anyway, could piss
reviewers off more than just omitting it

@@
 
-A more appropriate and useful place for extensions like this is directly within libraries, alongside abstractions that do not require this level of control.
-To enable this, the language must allow for the introduction new primitive type families, like \lstinline{Prod}, operators, like \lstinline{Pair}, \lstinline{PrL} and \lstinline{PrR}, and associated typechecking and translation logic, without allowing so much control that the reliability of the system suffers.
+A more appropriate and useful place for extensions like this is directly within libraries.
+To enable this, the language must allow for the introduction new primitive type families, like \lstinline{Prod}, operators, like \lstinline{Pair}, \lstinline{PrL} and \lstinline{PrR}, and associated typechecking and translation logic.
+% without allowing so much control that the reliability of the system suffers. 
 When encountering these new operators in expressions, the compiler must effectively hand control over typechecking and translation to the appropriate user-defined logic.

### 2013-10-12 esop14/att-esop14.tex
i removed a couple of sentences that are just circular
it'd be nice to have a better way to lead into the next section, but i
think the point is at least more clearly made this way.

@@
 The design described above suggests we may now need to add another layer to our language, an {extension language}, where extensions can be declared and implemented.
-In fact, we will show that \textbf{the most natural place for type system extensions is within the type-level language}.
-The intuition is that extensions to a language's static semantics will need to manipulate types as values at compile-time.
-Many languages already allow users to write type-level functions for various reasons, effectively supporting this notion of types as values at compile-time (see Sec.
+In fact, we will show that the most natural place for type system extensions is within the type-level language, with the intuition being that extensions to a language's static semantics will need to manipulate types as values at compile-time.
+
+% Many languages already allow users to write type-level functions, effectively supporting this notion of types as values at compile-time (see Sec.
 \ref{related-work} for examples).
@@
 This is precisely the structure that a distinct extension layer would have, and so we will demonstrate that it is quite natural to unify the two in this work.
+%
+% ^ the rest of this paragraph is a bit repetitive and circular.
+you wish to make the point that the type-level is the best place to include type system extensions because you have types as values at compile time.
+feel free to put it back, it just seems clearer to me this way.
 

### 2013-10-12 esop14/att-esop14.tex
Up through abs reps

--- 2013-10-12 esop14/att-esop14.tex
@@
 For example, regular expressions encoded using inductive datatypes are often considered overly verbose, so most functional languages support them via strings, which is less safe.
-Finally, general-purpose abstractions are implemented in a uniform manner, making it more difficult for domain knowledge to be applied to eliminate overhead or perform optimizations, leading to unsatisfactory performance in performance-critical applications -- particularly when targeting heterogeneous hardware platforms (e.g.
-programmable GPUs) and distributed computing resources.
-
-% Domain knowledge is not easily applied to eliminate overhead or perform optimizations, and implementations designed for typical application workloads may not be satisfactory in parts of a program where performance is a key criteria, particularly when targeting heterogeneous hardware platforms (e.g.
+Finally, general-purpose abstractions are implemented in a uniform manner.
+Domain knowledge is not easily applied to eliminate overhead or perform optimizations, and implementations designed for typical application workloads may not be satisfactory in parts of a program where performance is a key criteria, particularly when targeting heterogeneous hardware platforms (e.g.
 programmable GPUs) and distributed computing resources.
@@
 This discourages languages from including constructs that rely on statically-checked invariants stronger than those supported by their underlying implementation in the common intermediate language.
-At best, constructs like these can be exposed by generating a wrapper where run-time checks have been inserted to guarantee these invariants, thus compromising both verifiability and performance.
-%requires the development of an interoperability layer for every pair of DSLs. 
-%Moreover, this approach exposes the internals of an implementation to clients, making the abstraction awkward to work with and causing code breakage when implementation details change.
-This defeats a primary purpose of high-level programming languages: hiding low-level details from clients of an abstraction
+At best, constructs like these can be exposed by generating a wrapper where run-time checks have been inserted to guarantee these invariants.
+This compromises both verifiability and performance.
+For%requires the development of an interoperability layer for every pair of DSLs.
+Moreover, this approach exposes the internals of an implementation to clients, making the abstraction awkward to work with and causing code breakage when implementation details change.
+This defeats a primary purpose of high-level programming languages: hiding low-level details from clients of an abstraction.
 We diagram this fundamental \emph{interoperability problem} in Figure \ref{approaches}(a).
-%As an example, F\#'s type system prevents \lstinline{null} values from occurring within data structures, but because it's type system is not available when calling into F\# code from another language, like C\#, run-time null checks must still be included in the implementation.
+As an example, F\#'s type system prevents \lstinline{null} values from occurring within data structures, but because it's type system is not available when calling into F\# code from another language, like C\#, run-time null checks must still be included in the implementation.
 \begin{figure}[t]
@@
 Our choice of a simply-typed, simply-kinded calculus where expressions are given meaning by translation to a simply-typed internal language appears to occupy a ``sweet spot'' in the design space, and relates closely to how simply-typed functional languages like  ML and Haskell are specified and implemented today.
-%In Section \ref{design}, we briefly discuss other points in the design space of actively-typed languages and describe the sorts of abstractions that the mechanism as we have introduced it is not capable of expressing, suggesting several directions for future research.
+In Section \ref{design}, we briefly discuss other points in the design space of actively-typed languages and describe the sorts of abstractions that the mechanism as we have introduced it is not capable of expressing, suggesting several directions for future research.
 We conclude with a discussion of related work in Section \ref{related-work}.
-
 %dependently-typed and object-oriented type-level languages, as well as the constraints governing the design of the internal language.  
@@
 \section{From Extensible Compilers to Extensible Languages}\label{evolution}
-To understand the genesis of our internal extension mechanism, we begin by considering why most implementations of programming languages cannot even be  externally extended. 
+To understand the genesis of our internal extension mechanism, it is helpful to begin by considering why most implementations of programming languages cannot even be  externally extended. 
 Let us consider, as a simple example, an implementation of G\"odel's T, a typed lambda calculus with recursion on primitive natural numbers (see Appendix \ref{T}). 
@@
 If our newly-introduced constructs are exposed at a library's  interface boundary, clients using different compilers face the same problems with interoperability that those using different languages face.
-That is, extending a language by extending a single compiler for it is morally equivalent to creating a new language.
-Several prominent language ecosystems today are in a state where a prominent compiler has introduced or enabled the introduction of extensions that many libraries have come to rely on, including the Glasgow Haskell Compiler, SML/NJ and the GNU compilers for C and C++, thus complicating interoperability.
+That is, \textbf{extending a language by extending a single compiler for it is morally equivalent to creating a new language}.
+Several prominent language ecosystems today are in a state where a prominent compiler has introduced or enabled the introduction of extensions that many libraries have come to rely on, including the Glasgow Haskell Compiler, SML/NJ and the GNU compilers for C and C++.
+We argue that this practice should be considered harmful.
 
@@
 The result remains an integer according to the type system of the internal language so it is safe to give the entire expression type $\ttype{Nat}{\tunit}$ without violating the schema.
-Indeed, the extension mechanism will check that these consistencies hold (though not the kind system; see Sec.
+Indeed, the extension mechanism will check that these properties hold (see Sec.
 \ref{repcon}).
@@
 \subsection{Active Typechecking and Translation}
-The second phase of compilation involves typechecking and translating the program to produce an internal term, invoking the user-defined logic encoded in the introduced operator families when needed.
-The rules for active typechecking and translation are given in Fig.
-\ref{att}.
-The context $\fvalCtx$ tracks the operator definitions and representation schemas associated with families in scope.
-Because the logic inside operators must be invoked during this phase, we provide an evaluation semantics for type-level terms in Fig.
-\ref{tleval}.
-
-The key rule is \textsc{att-exp}, which assigns a type, $\tau$, and an \emph{abstract translation}, $\gabs$, to an expression, $e$, as specified by the judgement $\ecompilesX{e}{\tau}{\gabs}$, then \emph{deabstracts} the translation to produce the final translation.
-Deabstraction is specified in Fig.
-\ref{deabs} by the judgement $\eraseX{\gabs}{\gamma}$.
-The purpose of the abstract translation is to ensure that the implementation details of a type cannot be observed by operators associated with a different type, ensuring that invariants maintained by that type cannot be violated by operators in a different type family.
-For example, the implementation of natural numbers as integers in Fig.
-\ref{example} maintains the invariant that the translation is non-negative.
-If the knowledge that natural numbers are implemented as integers was externally visible, a different extension could introduce an operator like $\tops{badnum}{\kunit}{i}{a}{\tapp{\tvar{const}}{\tapp{\tvar{a}}{\tden{\titerm{-1}}{\ttype{Nat}{\tunit}}}}}$.
-
-This is prevented by a mechanism quite similar to the mechanism by which abstract types in ML-style modules operate \cite{atapl} -- by keeping the representation schema hidden outside the type family it is associated with.
-As a result, it cannot be shown given the knowledge available in the type family containing $\opvar{badnum}$ that the internal type associated with $\ttype{Nat}{\tunit}$ is $\dint$, and so $-1$ is not valid according to the rules we will describe below.
-The only way to produce a term of this type in an operator not associated with the type is to extract it from an input argument, where the invariants are guaranteed inductively to have been  maintained.
-We will make this more concrete in Sec.
-\ref{safety}.
-
-Our representation schema abstraction mechanism is specified syntactically, so it relates to previous work on syntactic type abstraction \cite{syntypeabs}.
-While that work was focused on abstracting away the identity of a particular type outside of a ``host'', we focus on abstracting away the knowledge of how a primitive type family is implemented outside of the operators associated with it.
-
-Variables translate directly to variables in the internal language, with the type determined by the typing context, $\Gamma$ (rule \textsc{att-var}).
-Lambda terms translate to lambda terms in the internal language.
-
-\subsection{Abstract Representations}
-Note that there is no elimination form that could be used to determine the internal type associated with a type, but the special internal type $\trepof{\tau}$, where $\tau$ is a type, can be used to refer to it abstractly
-
 \begin{figure}[t]
@@
 	\tEvalX{\tau_1}{\tau_1'}\\
-	\delfromtau{$\Xi_0$}{\fvalCtx}{\tau_1'}{\sabs}\\\\
+	%\delfromtau{$\Xi_0$}{\fvalCtx}{\tau_1'}{\sabs}\\\\
 	\ecompiles{\etCtxX{x}{\tau_1'}}{\fvalCtx}{e}{\tau_2}{\gabs}
 }{
-	\ecompiles{\etCtx}{\fvalCtx}{\elam{\evar{x}}{\tau_1}{e}}{\ttype{Arrow}{\tpair{\tau_1'}{\tau_2}}}{\ilam{\evar{x}}{\sabs}{\gabs}}
+	\ecompiles{\etCtx}{\fvalCtx}{\elam{\evar{x}}{\tau_1}{e}}{\ttype{Arrow}{\tpair{\tau_1'}{\tau_2}}}{\ilam{\evar{x}}{\trepof{\tau_1'}}{\gabs}}
 }
@@
 	\fvalDf \in \fvalCtx\\
-	\tops{op}{\kappai}{i}{a}{\tauop} \in \theta\\
-	\tEvalX{\taui}{\taui'}\\\\
+	\tops{op}{\kappaidx}{i}{a}{\tauop} \in \theta\\
+	\tEvalX{\tauidx}{\tauidx'}\\\\
 	\ecompiles{\etCtx}{\fvalCtx}{e_1}{\tau_1}{\gamma_1}~~~~
@@
 		\left[\text{$\begin{array}{r}
-			\taui'/\tvar{i}\\
+			\tauidx'/\tvar{i}\\
 			\tden{\titerm{\gamma_1}}{\tau_1} :: \cdots :: \tden{\titerm{\gamma_n}}{\tau_n} :: \tnil{\kDen}/\tvar{a}
@@
 		\Downarrow
-		\tden{\titerm{\gamma}}{\ttype{Fam'}{\tauidx}}
+		\tden{\titerm{\gabs}}{\tau}
 		\end{array}$}\\
-	\tauisdel{Fam}{\fvalCtx}{Fam'}{\tauidx}{\sigma}\\
+	\delfromtau{Fam}{\fvalCtx}{\tau}{\sabs}\\
 	\eCtxTogCtx{Fam}{\fvalCtx}{\etCtx}{\gtCtx}\\
-	\checkRC{\gtCtx}{Fam}{\fvalCtx}{\gamma}{\sigma}
-}{
-	\ecompiles{\etCtx}{\fvalCtx}{\eop{Fam}{op}{\taui}{\splat{e}{1}{n}}}{\ttype{Fam'}{\tauidx}}{\gamma}
-}
-\end{mathpar}
+	\checkRC{\gtCtx}{Fam}{\fvalCtx}{\gabs}{\sabs}
+}{
+	\ecompiles{\etCtx}{\fvalCtx}{\eop{Fam}{op}{\tauidx}{\splat{e}{1}{n}}}{\tau}{\gabs}
+}
+\end{mathpar}
+\vspace{-10pt}
 \caption{\small Active typechecking and translation of programs. Evaluation semantics for type-level terms are given in Fig. \ref{tleval}}
 \label{att}
+\vspace{-10pt}
 \end{figure}
+The second phase of compilation involves typechecking and translating the program to produce an internal term, invoking the user-defined logic implemented in operator families when needed.
+The rules for this phase are given in Fig.
+\ref{att}.
+The context $\fvalCtx$ tracks operator definitions and representation schemas associated with families in scope.
+Because the logic inside operators must be invoked during this phase, we need an evaluation semantics for type-level terms, given in Fig.
+\ref{tleval}.
+This is a largely unsurprisingly collection of rules with the exception of \textsc{dencase-eval-valid}, which we will explain below.
+
+\subsection{Abstract Translations and Abstract Internal Types}
+The rule \textsc{att-exp} first assigns a type, $\tau$, and an \emph{abstract translation}, $\gabs$, to an expression, $e$, as determined by the judgement $\ecompilesX{e}{\tau}{\gabs}$, and then \emph{deabstracts} the abstract translation to complete compilation.
+Deabstraction is specified in Fig.
+\ref{deabs} by the judgement $\eraseX{\gabs}{\gamma}$.
+The purpose of the abstract translation is to ensure that the implementation details of a type family are not exposed to operators associated with a different type family, so that invariants are preserved when families are composed.
+For example, our implementation of natural numbers as integers in Fig.
+\ref{example} maintains the invariant that the translation is non-negative.
+If the knowledge that natural numbers are implemented as integers was externally visible, a different extension could introduce an operator like $\tops{badnat}{\kunit}{i}{a}{\tapp{\tvar{const}}{\tapp{\tvar{a}}{\tden{\titerm{-1}}{\ttype{Nat}{\tunit}}}}}$, breaking this invariant (and thus the guarantee that the recursor always terminates!)
+%To put it another way, if a collection of operators can be shown to be a full and faithful (that is, adequate) encoding of a particular type system, then it does not matter what other types there are in the system.
+
+We prevent this by a mechanism quite similar to the mechanism by which abstract types in ML-style modules operate \cite{atapl}: by keeping the representation schema hidden outside the type family it is associated with.
+Thus, it cannot be shown given the knowledge available in the type family containing $\opvar{badnum}$ that the internal type associated with $\ttype{Nat}{\tunit}$ is $\dint$, and thus $-1$ is not valid according to the rules we will describe below.
+The only way to produce a term of type $\ttype{Nat}{\tunit}$ as the result of applying an operator not associated with the family $\fvar{Nat}$ is if it extracts a previously-constructed value of that type  out of an argument (e.g.
+when projecting it out of an $n$-tuple), in which case the invariants are inductively  maintained.
+%
+
+Let us begin by going over how expressions are assigned types and abstract translations.
+Variables (\textsc{att-var}) translate directly to variables in the internal language, with the type determined by the typing context, $\Gamma$.
+Lambda terms (\textsc{att-lam}) also translate to lambda terms in the internal language and are given the appropriate $\fvar{Arrow}$ type.
+The internal type of the argument is $\trepof{\tau_1'}$, which is called the \emph{abstract internal type} associated with $\tau_1'$ (which is the argument type $\tau_1$ after it has been fully evaluated to a type-level value).
+This will be converted to the actual internal type determined by $\tau'$'s representation schema during deabstraction.
+Otherwise, typechecking and translation proceeds recursively be extending the typing context as is usual.
+
+The most interesting case is operator application (\textsc{att-op}).
+This rule extracts the appropriate operator definition, evaluates the operator index to a value, $\tauidx'$, and then recursively assigns a type and abstract translation to each argument.
+Next, it constructs a list of denotations from these and passes it, along with the fully evaluated operator index, to the implementation of the operator, $\tauop$.
+If this evaluates to a valid denotation, then compilation can proceed.
+Otherwise, compilation will stop, here because no other rule will apply, but in practice by displaying a type error for the user.
+
+The resulting valid denotation cannot be used directly, however.
+First, we must check that it is \emph{representationally consistent}, meaning that the abstract internal term produced, $\gabs$, can be shown to have the abstract internal type associated with the type, $\tau$, that was assigned to the expression.
+The judgement $\delfromtau{Fam}{\fvalCtx}{\tau}{\sabs}$ determines the abstract internal type, and $\checkRC{\gtCtx}{Fam}{\fvalCtx}{\gabs}{\sabs}$ performs the check.
+To do so, the variable context $\Gamma$, which associates variables with types, must be converted to a context associating variables with their corresponding abstract internal type, $\gtCtx$, as specified by the judgement $\eCtxTogCtx{Fam}{\fvalCtx}{\etCtx}{\gtCtx}$.
+Each of these judgements are defined in Fig.
+\ref{ait}.
+
+To make this fundamental mechanism clear, and explain the rule \textsc{dencase-eval-valid} mentioned above, let us go through how the successor operator from Fig.
+\ref{example} works during this phase.
+
+The special internal term $\tvalof{\titerm{\gamma}}{\tautype}$ produced when the rule $\textsc{dencase-eval-valid}$ is applied is what allows the actual term produced by an operator outside of the local family to be held abstract.
+
 \begin{figure}[t]
@@
 
-\inferrule[dencase eval valid]{
+\inferrule[dencase-eval-valid]{
 	\tEvalX{\tau}{\tden{\tauiterm}{\tautype}}\\\\
@@
 \section{Related Work}\label{related-work}
+Our representation schema abstraction mechanism is purely syntactic, so it relates to previous work on syntactic type abstraction \cite{syntypeabs}.
+While that work was focused on abstracting away the identity of a particular type outside of the ``principal'' it is associated with, we focus on abstracting away the knowledge of how a primitive type family is implemented outside of its operators.
+
+Compiler correctness / ML work (Harper, Crary).
+
+Erdweg's work (see email)
+
+Term rewriting systems
+
 \subsection{Type-Level Computation} %Haskell, Ur and $\Omega$mega
@@
 
-\subsection{Run-Time Indirection}
-{\it Operator overloading} \cite{vanWijngaarden:Mailloux:Peck:Koster:Sintzoff:Lindsey:Meertens:Fisker:acta:1975} and {\it metaobject dispatch} \cite{Kiczales91} are run-time protocols that translate operator invocations into function calls.
+% \subsection{Run-Time Indirection}
+% {\it Operator overloading} \cite{vanWijngaarden:Mailloux:Peck:Koster:Sintzoff:Lindsey:Meertens:Fisker:acta:1975} and {\it metaobject dispatch} \cite{Kiczales91} are run-time protocols that translate operator invocations into function calls.
 The function is typically selected according to the type or value of one or more operands.

### 2013-10-12 esop14/att-esop14.tex

@@
 If our newly-introduced constructs are exposed at a library's interface boundary, clients using different compilers face the same problems with interoperability that those using different languages face.
-That is, \textbf{extending a language by extending a single compiler for it is morally equivalent to creating a new language}.
-Several prominent language ecosystems today are in a state where a prominent compiler has introduced or enabled the introduction of extensions that many libraries have come to rely on, including the Glasgow Haskell Compiler, SML/NJ and the GNU compilers for C and C++.
-We argue that this practice should be considered harmful.
-
-A more appropriate and useful place for extensions like this is directly within libraries, alongside abstractions that do not require this level of control.
-To enable this, the language must allow for the introduction new primitive type families, like \lstinline{Prod}, operators, like \lstinline{Pair}, \lstinline{PrL} and \lstinline{PrR}, and associated typechecking and translation logic, without allowing so much control that the reliability of the system suffers.
+That is, extending a language by extending a single compiler for it is morally equivalent to creating a new language.
+Several prominent language ecosystems today are in a state where a prominent compiler has introduced or enabled the introduction of extensions that many libraries have come to rely on, including the Glasgow Haskell Compiler, SML/NJ and the GNU compilers for C and C++, thus complicating interoperability.
+
+A more appropriate and useful place for extensions like this is directly within libraries.
+To enable this, the language must allow for the introduction new primitive type families, like \lstinline{Prod}, operators, like \lstinline{Pair}, \lstinline{PrL} and \lstinline{PrR}, and associated typechecking and translation logic.
+% without allowing so much control that the reliability of the system suffers. 
 When encountering these new operators in expressions, the compiler must effectively hand control over typechecking and translation to the appropriate user-defined logic.

### 2013-10-12 esop14/att-esop14.tex
ESOP submission

--- 2013-10-12 esop14/att-esop14.tex
@@
 This approach is often effective because modern {general-purpose} abstraction mechanisms, like inductive datatypes and object systems, are highly expressive.  
-For example, the Delite framework leverages Scala's powerful general-purpose mechanisms to enable a variety of interesting \emph{embedded domain-specific languages} \cite{delite}. 
+For example, the Delite framework leverages Scala's powerful general-purpose mechanisms to enable a number of useful \emph{embedded domain-specific languages}.
+%\cite{delite}
 Unfortunately, there remain some situations of interest where general-purpose abstractions fall short. 
 For instance, it is difficult to adequately encode advanced type systems in terms of the simpler rules governing general-purpose abstractions (e.g.
-reasoning about units of measure requires built-in language support in F\# \cite{units-of-measure}).
+reasoning about units of measure requires built-in language support in F\# \cite{conf/cefp/Kennedy09}).
 Even if a full encoding is possible, it may not be useful if it is overly verbose or unnatural, or if the error messages are overly abstract.
@@
 programmable GPUs) and distributed computing resources.
-Using variants of familiar abstractions backed by specialized implementation strategies are quite useful in these cases.
+Using variants of familiar abstractions backed by specialized implementation strategies can be quite useful in these cases.
 
@@
 We diagram this fundamental \emph{interoperability problem} in Figure \ref{approaches}(a).
-As an example, F\#'s type system prevents \lstinline{null} values from occurring within data structures, but because it's type system is not available when calling into F\# code from another language, like C\#, run-time null checks must still be included in the implementation.
+%As an example, F\#'s type system prevents \lstinline{null} values from occurring within data structures, but because it's type system is not available when calling into F\# code from another language, like C\#, run-time null checks must still be included in the implementation.
 \begin{figure}[t]
@@
 To make this proposal concrete, we begin by introducing a simple core calculus, called \atlam~(for the ``actively-typed lambda calculus''), in Section \ref{atlam}. 
-This calculus uses type-level computation of higher kind, along with techniques borrowed from the typed compilation literature and a form of type abstraction that ensures that the implementation details of an extension are not externally visible to guarantee the safety of the language, the decidability of typechecking and compilation and composability of extensions.
-
-In Section \ref{examples}, we show that despite these constraints, this mechanism is expressive enough to admit, within libraries, a number of general-purpose and domain-specific abstractions that normally require built-in language support.
-Our core calculus uses a uniform abstract syntax for primitive operators to simplify our presentation and analysis, but this syntax is too verbose to be practical.
+This calculus uses type-level computation of higher kind, along with techniques borrowed from the typed compilation literature and a form of type abstraction that ensures that the implementation details of an extension are not externally visible to guarantee the safety of the language, the decidability of typechecking and compilation and composability of extensions, as we outline in Section \ref{safety}.
+In Section \ref{examples}, we suggest that despite these constraints, this mechanism is expressive enough to admit, within libraries, a number of general-purpose and domain-specific abstractions that normally require built-in language support.
+
+%Our core calculus uses a uniform abstract syntax for primitive operators to simplify our presentation and analysis, but this syntax is too verbose to be practical.
 Thus, we begin this section by showing how a key design choice made in the calculus -- to associate operators with type families, forming what we call \emph{active type families} -- supports a novel type-directed desugaring mechanism that permits the use of conventional concrete syntax for language extensions.
 
-Our choice of a simply-typed, simply-kinded calculus where expressions are given meaning by translation to a simply-typed internal language appears to occupy a ``sweet spot'' in the design space, and relates closely to how simply-typed functional languages like  ML and Haskell are specified and implemented today.
+% Our choice of a simply-typed, simply-kinded calculus where expressions are given meaning by translation to a simply-typed internal language appears to occupy a ``sweet spot'' in the design space, and relates closely to how simply-typed functional languages like  ML and Haskell are specified and implemented today.
 In Section \ref{design}, we briefly discuss other points in the design space of actively-typed languages and describe the sorts of abstractions that the mechanism as we have introduced it is not capable of expressing, suggesting several directions for future research.
@@
 To understand the genesis of our internal extension mechanism, it is helpful to begin by considering why most implementations of programming languages cannot even be  externally extended. 
-Let us consider, as a simple example, an implementation of G\"odel's T, a typed lambda calculus with recursion on primitive natural numbers (see Appendix \ref{T}). 
+Let us consider, as a simple example, an implementation of G\"odel's T, a typed lambda calculus with recursion on primitive natural numbers (see Appendix). 
 A compiler for this language written using a functional language will invariably represent the primitive type families and operators using {closed} inductive datatypes. 
@@
 In an object-oriented implementation of Godel's T, we might instead encode types and operators as subclasses of abstract classes \lstinline{Type} and \lstinline{Exp}.
-Typechecking and translation will proceed by the ubiquitous \emph{visitor pattern} \cite{visitor} by dispatching against a fixed collection of {known} subclasses of \lstinline{Exp}.
+Typechecking and translation will proceed by the ubiquitous \emph{visitor pattern}  by dispatching against a fixed collection of {known} subclasses of \lstinline{Exp}.
 
 In either case, we encounter the same basic issue: there is no way to modularly add new primitive type families and operators and implement their associated typechecking and translation logic. 
-This issue is related to the widely-discussed \emph{expression problem} (in a restricted sense -- we do not consider adding new functions beyond typechecking and translation here, only adding logic to these) \cite{wadler-expression}.
+%This issue is related to the widely-discussed \emph{expression problem} (in a restricted sense -- we do not consider adding new functions beyond typechecking and translation here, only adding logic to these) \cite{wadler-expression}.
 
 A number of language mechanisms have been proposed that allow new cases to be added to datatypes and the functions that operate over them in a modular manner. 
-In functional languages, we might use \emph{open datatypes} \cite{open-datatypes}.
+In functional languages, we might use \emph{open datatypes}.
 For example, if we wish to extend G\"odel's T with product types and we have written our compiler in a language supporting open inductive datatypes, it might be possible to add new cases like this:
@@
 If our newly-introduced constructs are exposed at a library's  interface boundary, clients using different compilers face the same problems with interoperability that those using different languages face.
-That is, \textbf{extending a language by extending a single compiler for it is morally equivalent to creating a new language}.
+That is, {extending a language by extending a single compiler for it is morally equivalent to creating a new language}.
 Several prominent language ecosystems today are in a state where a prominent compiler has introduced or enabled the introduction of extensions that many libraries have come to rely on, including the Glasgow Haskell Compiler, SML/NJ and the GNU compilers for C and C++.
-We argue that this practice should be considered harmful.
 
@@
 The type-level language is often constrained by its own type system (where the types of type-level values are called \emph{kinds} for clarity) that prevents type-level functions from causing problems during compilation.
-This is precisely the structure that a distinct extension layer would have, and so we will demonstrate that it is quite natural to unify the two in this work.
+This is precisely the structure that a distinct extension layer would have, and so it is quite natural to unify the two.
 
@@
 \end{figure}
-In this section, we will develop a core calculus, called @$\lambda$ for the ``actively-typed lambda calculus'', by way of a semantics and a simple example, and discuss how it addresses the safety concerns that arise when giving users this level of control over the semantics of a language and its implementation.
+In this section, we will develop a core calculus, called @$\lambda$ for the ``actively-typed lambda calculus'', by way of a semantics and a simple example, and discuss how it addresses the safety concerns that arise.
 \subsection{Overview}
-The grammar of \atlam~is shown in Figure \ref{grammar}. The language is a simply-typed lambda calculus with simply-kinded type-level computation. Kinds, $\kappa$, classify type-level terms, $\tau$. 
-Types are type-level values of kind $\star$ (following System $F_{\omega}$ \cite{fomega}) and classify expressions, $e$.
-The type-level language also includes other kinds of terms, such as type-level functions, lists (required by our mechanism) and integers, strings and products for the sake of our examples (see Sec.
-\ref{termination} for a discussion on other acceptable kinds of type-level data).
-It also includes constructs for developing extensions -- denotations and reified internal terms and types -- which we will discuss in the sections below.
-
-At the top level, programs, $\rho$, consist of a series of declarations followed by an expression.
+The grammar of \atlam~is shown in Figure \ref{grammar}. 
+A program, $\rho$, consists of a series of declarations followed by an expression.
 Declarations can be either bindings of type-level terms to type-level variables using \textsf{def} or a declaration of a new primitive type family using \textsf{family}.
-Expressions can be either variables, lambdas, or applications of operators, and are ultimately given meaning by translation to a typed internal language.
+Expressions, $e$, can be either variables, lambdas, or applications of operators, and are ultimately given meaning by translation to a typed internal language, with terms $\gamma$ and types $\sigma$.
 This language has been chosen, for simplicity, to be a variant of Plotkin's PCF with primitive integers and products, but in practice would include other constructs consistent with its role as a high-level intermediate language.
-The grammars of internal terms, $\gamma$, and internal types, $\sigma$, also include special forms containing type-level terms; these are used for developing extensions and during compilation and will be erased before compilation ends, as we will explain below.
-
-\subsection{Example: G\"odel's T as an Active Type Family}
+
+The language is structured as a simply-typed lambda calculus with simply-kinded type-level computation. Kinds, $\kappa$, classify type-level terms, $\tau$. 
+Types are type-level values of kind $\star$ (following System $F_{\omega}$) and classify expressions.
+The type-level language also includes other kinds of terms: type-level functions, lists (required by our mechanism), integers, strings and products for the sake of our examples (see Sec.
+\ref{termination} for a discussion on other acceptable kinds of type-level data) and constructs for developing extensions -- denotations and reified internal terms and types -- which we will discuss in the sections below.
+The grammar of internal terms, $\gamma$, and internal types, $\sigma$, also includes special forms containing type-level terms; these are used for developing extensions and during compilation and will be erased before compilation ends.
 \begin{figure}[t]
@@
 \label{z}&\quad\tops{z}{\kunit}{i}{a}{
-	\tapp{\tapp{\tvar{const}}{\tvar{a}}}{
+	\tapp{\tapp{\tvar{empty}}{\tvar{a}}}{
 		\tden{\titerm{0}}{\ttype{Nat}{\tunit}}
@@
 \caption{G\"odel's T in \atlam, used to calculate 2+2.
-The statics we implement here are shown in Appendix \ref{T}.
-Simple helper functions for working with argument lists ($\tvar{const}$, $\tvar{pop}$, $\tvar{pop\_final}$) and types ($\tvar{check\_type}$) are defined in Appendix \ref{helper}.}
+The statics we implement here are shown in the Appendix.
+Simple helper functions for working with lists ($\tvar{empty}$, $\tvar{pop}$, $\tvar{pop\_final}$) and types ($\tvar{check\_type}$),  described below, are given in the appendix.}
 \label{example}
-\vspace{-10pt}
 \end{figure}
+\subsection{Example: G\"odel's T as an Active Type Family}
 To make our explanation of each of the constructs in the calculus concrete, we will work through an example showing how to introduce primitive natural numbers with bounded recursion in the style of G\"odel's T \cite{pfpl}.
-These will be implemented internally as integers (internal terms of internal type $\dint$).
-Figure \ref{example} shows how the indexed type family $\fvar{Nat}$ is defined.
-This family is indexed by a unit value (of kind $\kunit$), so it contains only one type, $\ttype{Nat}{\tunit}$, which we alias on line \ref{nattype} by defining the type-level variable $\tvar{nat}$.
+These will be implemented internally as integers (that is, internal terms of internal type $\dint$).
+Figure \ref{example} shows how to define the indexed type family $\fvar{Nat}$.
+This family contains only one type, written $\ttype{Nat}{\tunit}$, which we alias on line \ref{nattype} by defining the type-level variable $\tvar{nat}$.
 We define the typechecking and translation logic for the operators associated with this family (\opvar{z}, \opvar{s} and \opvar{rec}) on lines \ref{z}-\ref{lastop} and use these to define a $plus$ function on line \ref{add} and compute 2+2.
@@
 The syntactic form $\familyDf$ declares a new primitive type family named $\fvar{Fam}$ indexed by type-level values of kind $\kappaidx$ with representation schema $\tvar{i}.\taurep$ and operators $\theta$.
-This can be compared to adding a new constructor to the compiler-internal datatype \lstinline{Type}, as suggested in Sec.
-\ref{evolution}.
-The index represents the data associated with the constructor.
+The purpose of the representation schema and of associating of operators directly with types will be explained below.
+
+Declaring a type family in this way is a language-internal analog to adding a new constructor to the compiler-internal datatype \lstinline{Type}, as suggested in Sec. \ref{evolution}. 
+The index represents the data associated with this constructor.
+A type (that is, a type-level term of kind $\kTypeBlur$) is constructed by naming a family in scope and providing a type-level term of the appropriate kind as an index.
 A base type like $\tvar{nat}$ can be thought of as being the only type in the family $\fvar{Nat}$ trivially indexed by the unit value, of kind $\kunit$, while families like $\fvar{Ntuple}$ might be indexed by a list of types, having kind $\klist{\kTypeBlur}$.
-A type (that is, a type-level term of kind $\kTypeBlur$) is constructed by naming a family in scope and providing a type-level term of the appropriate kind as an index.
  For example, $\ttype{Ntuple}{\tcons{\tvar{nat}}{\tcons{\tvar{nat}}{\tnil{\kTypeBlur}}}}$ might be the type of a pair of natural numbers.
+Given a type, its family can be \textsf{case} analyzed to extract the value of its index.
 It is important that type equality be decidable, so only kinds for which equivalence coincides with syntactic equality can be used as type family indices.
 The main  consequence of this restriction is that indices cannot contain type-level functions.
-Given a type, its family can be \textsf{case} analyzed to extract its index.
-
-\subsection{Representation Schemas}
-As we will discuss further below, it is important that all expressions classified by a type compile to consistently-typed internal terms\todo{mention compiler correctness history here?}.
-For this reason, we require that every type have associated with it a single internal type.
-This is computed by substituting the type index for the bound variable $\tvar{i}$ in the term $\taurep$, called the \emph{representation schema}, associated with the type family and evaluating to a value representing an internal type.
-Internal types, $\sigma$, are reified into type-level terms of kind $\kIType$ using the introductory form $\titype{\sigma}$.
+
+\subsection{Representations and Representation Schemas}
+As we will discuss further below, it is important that all expressions classified by a type compile to consistently-typed internal terms.
+For this reason, we require that every type have a single internal type associated with it, called its \emph{representation}.
+This is computed by substituting the type index for the bound variable $\tvar{i}$ in the term $\taurep$, called the \emph{representation schema} of the type family, and evaluating to a value representing an internal type.
+Internal types, $\sigma$, are reified as type-level terms of kind $\kIType$ using the introductory form $\titype{\sigma}$.
 
 \subsection{Indexed Operator Families and Denotations}\label{operators}
-Type families also have a collection of primitive operator families associated with them.
+Type families are also equiped with a collection of primitive operator families.
 An operator family named $\opvar{op}$ is declared using the form $\tops{op}{\kappaidx}{i}{a}{\tauop}$.
-Like type families, operator families are indexed by values of some kind, $\kappaidx$.
-Operators are not first-class type-level values in our calculus, so there are no restrictions on this kind related to equality.
+Like type families, operator families are indexed by values of some kind, $\kappaidx$, but because operators are not first-class type-level values in our calculus, there are no equality restrictions.
 In the example in Fig.
@@
 an integer in our calculus).
-A family implementing record types or object types might have a similar operator indexed by a type-level string representing the field being accessed.
-
-To select an operator from a family by providing an index and apply it to $n$ arguments, the grammar provides a uniform form: $\eop{Fam}{op}{\tauidx}{e_1; \ldots; e_n}$, where $n \geq 0$.
-The typechecking and translation of an expression of this form is controlled by the implementation, $\tauop$.
-It must evaluate to a \emph{denotation}, which is a type-level value of kind $\kDen$, when given the operator index, $\tauidx$, and a list constructed from the denotations recursively assigned to each argument, $e_1$ through $e_n$.
-There are two forms of denotations that an expression can have.
-A \emph{valid denotation} has the form $\tden{\tauiterm}{\tautype}$, where $\tauiterm$ represents the \emph{translation} of the expression to an internal term and $\tautype$ is the type assigned to it.
-Internal terms are reified as type-level terms using the form $\titerm{\gamma}$, and have kind $\kITerm$ (similar to $\kIType$).
-It is important to note that there are no elimination form for reified terms or types.
+A family implementing record types or object types might have a similar operator indexed by a type-level string representing the field being accessed (see Sec.
+\ref{examples}).
+
+To apply an operator, the grammar provides a uniform form of expression: $\eop{Fam}{op}{\tauidx}{e_1; \ldots; e_n}$, where $n \geq 0$.
+The typechecking and translation of an expression of this form is controlled by the term $\tauop$ in the operator's declaration.
+This term must evaluate to a \emph{denotation}, which is a type-level value of kind $\kDen$, when given the operator index, $\tauidx$, and a list constructed from the denotations recursively assigned to each argument, $e_1$ through $e_n$.
+
+There are two forms of denotations that an expression can be assigned.
+A \emph{valid denotation} has the form $\tden{\tauiterm}{\tautype}$, where $\tauiterm$ is the \emph{translation} of the expression to an internal term and $\tautype$ is the type it has been assigned.
+Internal terms are represented as type-level terms using the form $\titerm{\gamma}$, and have kind $\kITerm$ (similar to $\kIType$).
+%It is important to note that there are no elimination form for reified terms or types.
 %(indeed, in \atlam, terms are never  syntactically analyzed directly by extensions, unlike macro systems and other forms of term rewriting; see Sec. \ref{related-work})
-If a type cannot be assigned to an operation according to the specification of the operator being implemented, then the \emph{error denotation}, $\terr$, is returned instead of a valid denotation.
-In a more practical implementation, a specialized error message and other diagnostic information could be associated with $\terr$, but we omit such details for the sake of simplicity in our core calculus.
+If a type error is detected by an operator, the \emph{error denotation}, $\terr$, is returned instead of a valid denotation.
+In a practical implementation, a specialized error message and other diagnostic information would be provided when returning $\terr$, but we omit such details for simplicity.
 Terms of kind $\kDen$ can be \textsf{case} analyzed to determine if they are valid or errors, and if valid, to extract the translation and type.
@@
 In the example in Fig.
-\ref{example}, the operator $\opvar{z}$ checks that no arguments were provided using the simple helper function $\tvar{const}$ (shown in Appendix \ref{helper}).
-If so, it returns a valid denotation by pairing the translation $\titerm{0}$ with the type $\ttype{Nat}{\tunit}$, as expected\footnote{In fact, there is no theoretical barrier to a different ``zero'' being used, since integers and natural numbers are both of countably infinite cardinality!}.
-If not, the helper function returns $\terr$.
-The successor operator, $\opvar{s}$, does take an argument, so it pops a denotation off the argument list, making sure there are no more, and binds its translation and type to $\tvar{x}$ and $\tvar{t}$ respectively, all using the helper function $\tvar{pop\_final}$.
+\ref{example}, the operator $\opvar{z}$ checks that no arguments were passed in using the simple helper function $\tvar{empty}$.
+If so, it returns a valid denotation by pairing the translation $\titerm{0}$ with the type $\ttype{Nat}{\tunit}$, as expected\footnote{Actually, there is no theoretical barrier to a different ``zero'' being used!}.
+If an argument was provided, the helper function returns $\terr$.
+The successor operator, $\opvar{s}$, takes one argument, so it pops a denotation off the argument list, making sure there are no more, and binds its translation and type to $\tvar{x}$ and $\tvar{t}$ respectively, all using the helper function $\tvar{pop\_final}$.
 It then checks that the argument's type is also $\ttype{Nat}{\tunit}$, returning a denotation pairing the translation $\titerm{\iup{\tvar{x}} + 1}$ with the type $\ttype{Nat}{\tunit}$ if so.
-The internal language includes a special form $\iup{\tau}$ which is used to ``un-reify'' reified internal terms, of kind $\kITerm$ (thus serving as the left-inverse of $\titerm{\gamma}$).
-In this case, $\tvar{x}$ is a type-level variable of kind $\kITerm$ representing the translation of the argument to the successor operator.
-Because we have checked that it is a natural number, and natural numbers always translate to integers by the representation schema, we know that it is safe to add 1 to this internal term.
-The result remains an integer according to the type system of the internal language so it is safe to give the entire expression type $\ttype{Nat}{\tunit}$ without violating the schema.
-Indeed, the extension mechanism will check that these properties hold (see Sec.
-\ref{repcon}).
+The form $\iup{\tau}$ is used to ``un-reify'' reified internal terms, of kind $\kITerm$ (thus serving as the left-inverse of $\titerm{\gamma}$).
+In this case, $\tvar{x}$ is a type-level variable of kind $\kITerm$ representing the translation of the argument to the successor operator, so we simply need to add one to it.
+Because we have checked that the denotation's type was $\ttype{Nat}{\tunit}$, and the representation schema will guarantee that expressions of this type always translate to integers, we know that it is safe to do so.
 If any of these steps fail, the various helper functions we use simply return $\terr$ (in practice, it would be prudent to equip each failure condition with a different error message).
+Compilation will also fail if we accidentally violate the representation schema, as we will see in Sec.
+\ref{repcon}.
 
@@
 \subsection{Functions, Variables and Arrow Types}
+
 The recursor operator, $\opvar{rec}$, proceeds similarly, extracting the translations and types of each of its three arguments.
-The interesting part comes when analyzing the third argument, the recursive case, which binds two variables (the predecessor and the result of recursing on it).
+Of note, however, is how it handles the third argument, which binds two variables (the predecessor and the result of recursing on it).
 In \atlam, the built-in $\lambda$ operator serves as the sole mechanism for introducing bound variables.
-In most calculi, lambda terms have types of the form $\tau_1 \rightarrow \tau_2$, which in \atlam~corresponds to a built-in type family, $\fvar{Arrow}$, indexed by a pair of types, $\kpair{\kTypeBlur}{\kTypeBlur}$, that is always in scope.
+In most calculi, lambda terms have types of the form $\tau_1 \rightarrow \tau_2$.
+In \atlam, $\rightarrow$ corresponds to the built-in type family, $\fvar{Arrow}$.
+This family is indexed by a pair of types, $\kpair{\kTypeBlur}{\kTypeBlur}$, and is always in scope.
 Its representation schema simply maps to the corresponding internal arrow type, $\tvar{i}.\titype{\darrow{\trepof{\tfst{\tvar{i}}}}{\trepof{\tsnd{\tvar{i}}}}}$.
-The special form $\trepof{\tau}$ is used to refer, abstractly, to the representation of the two types it is indexed by (we also see this used on line \ref{fix}).
-Although $\lambda$ is built-in, application is an operator associated with the $\fvar{Arrow}$ family, $\opvar{ap}$, as is seen on line \ref{ap}.
-\todo{figure for arrow types}
+The special form $\trepof{\tau}$ is used to refer, abstractly, to the representation of the type $\tau$ (we also see this used on line \ref{fix}).
+Although the $\lambda$ operator is built-in, because it needs to bind variables, application is just an operator associated with the $\fvar{Arrow}$ family, $\opvar{ap}$, as is seen on line \ref{ap}.
+The details are straightforward and given in the appendix.
 
@@
 The kinding rules for programs are given in Figure \ref{kindprog}, and they rely on the kinding rules for type-level terms given in Figure \ref{tlkind}.
-These rules use contexts $\Sigma$ and $\Theta$ to track family and operator signatures, but beyond that, the rules are largely consistent with those of a simply-typed lambda calculus, with the addition of the handful of special forms constrained as described in the previous sections.
+These rules use contexts $\Sigma$ and $\Theta$ to track family and operator signatures, but beyond that, the rules are largely consistent with those of a simply-typed lambda calculus shifted into the type-level, with the addition of the handful of special forms constrained as described in the previous sections.
 The reader is encouraged to verify that the example in Fig.
@@
 \label{kindprog}
+\vspace{-10pt}
 \end{figure}
-\subsection{Active Typechecking and Translation}
+%\subsection{Active Typechecking and Translation}
 \begin{figure}[t]
@@
 		\end{array}$}\\
-	\delfromtau{Fam}{\fvalCtx}{\tau}{\sabs}\\
-	\eCtxTogCtx{Fam}{\fvalCtx}{\etCtx}{\gtCtx}\\
-	\checkRC{\gtCtx}{Fam}{\fvalCtx}{\gabs}{\sabs}
+	%\delfromtau{Fam}{\fvalCtx}{\tau}{\sabs}\\
+	%
+	\ddbar{\Xi_0,\fvar{Fam}}{\fvalCtx}{\trepof{\tau}}{\sabs}\\
+	\eCtxTogCtx{\Xi_0,\fvar{Fam}}{\fvalCtx}{\etCtx}{\gtCtx}\\
+	\checkRC{\gtCtx}{\Xi_0,\fvar{Fam}}{\fvalCtx}{\gabs}{\sabs}
 }{
@@
 \vspace{-10pt}
-\caption{\small Active typechecking and translation of programs. Evaluation semantics for type-level terms are given in Fig. \ref{tleval}}
+\caption{\small Active typechecking and translation of programs. Evaluation semantics for type-level terms are given in Fig. \ref{tleval}.}
 \label{att}
@@
 \end{figure}
-The second phase of compilation involves typechecking and translating the program to produce an internal term, invoking the user-defined logic implemented in operator families when needed.
+The second phase of compilation involves invoking the  logic implemented by user-defined operator families to typecheck and translate the program.
 The rules for this phase are given in Fig.
 \ref{att}.
-The context $\fvalCtx$ tracks operator definitions and representation schemas associated with families in scope.
+The context $\fvalCtx$ tracks the operator definitions and representation schemas associated with families in scope.
 Because the logic inside operators must be invoked during this phase, we need an evaluation semantics for type-level terms, given in Fig.
 \ref{tleval}.
-This is a largely unsurprisingly collection of rules with the exception of \textsc{dencase-eval-valid}, which we will explain below.
-
-\subsection{Abstract Translations and Abstract Internal Types}
-The rule \textsc{att-exp} first assigns a type, $\tau$, and an \emph{abstract translation}, $\gabs$, to an expression, $e$, as determined by the judgement $\ecompilesX{e}{\tau}{\gabs}$, and then \emph{deabstracts} the abstract translation to complete compilation.
-Deabstraction is specified in Fig.
-\ref{deabs} by the judgement $\eraseX{\gabs}{\gamma}$.
-The purpose of the abstract translation is to ensure that the implementation details of a type family are not exposed to operators associated with a different type family, so that invariants are preserved when families are composed.
+This is again a largely unsurprising collection of rules with the exception of \textsc{dencase-eval-valid}, which we will explain below.
+
+\subsection{Abstract Representations and Abstract Internal Types}\label{repcon}
+To typecheck and translate an expression, $e$, the rule \textsc{att-exp} first assigns a type, $\tau$, and an \emph{abstract translation}, $\gabs$, to it, as determined by the judgement $\ecompilesX{e}{\tau}{\gabs}$.
+It then \emph{deabstracts} this abstract translation to complete the compilation process, as determined by the judgement $\eraseX{\gabs}{\gamma}$.
+The purpose of the abstract translation phase is to ensure that the implementation details of type families are not exposed to other families, so that invariants that they rely on are preserved when families are composed.
 For example, our implementation of natural numbers as integers in Fig.
@@
 
-We prevent this by a mechanism quite similar to the mechanism by which abstract types in ML-style modules operate \cite{atapl}: by keeping the representation schema hidden outside the type family it is associated with.
-Thus, it cannot be shown given the knowledge available in the type family containing $\opvar{badnum}$ that the internal type associated with $\ttype{Nat}{\tunit}$ is $\dint$, and thus $-1$ is not valid according to the rules we will describe below.
-The only way to produce a term of type $\ttype{Nat}{\tunit}$ as the result of applying an operator not associated with the family $\fvar{Nat}$ is if it extracts a previously-constructed value of that type  out of an argument (e.g.
-when projecting it out of an $n$-tuple), in which case the invariants are inductively  maintained.
-%
-
-Let us begin by going over how expressions are assigned types and abstract translations.
+We preclude such operations by a mechanism similar to the abstract type mechanism supported by ML-style module systems \cite{pfpl}.
+By keeping a type family's representation schema private to the operators associated with it, they maintain full control over representation invariants.
+So, because it cannot be shown given the knowledge available in the type family containing $\opvar{badnum}$ that the internal type associated with $\ttype{Nat}{\tunit}$ is $\dint$, the translation $-1$ will not be permitted according to the rules we will give below.
+The only way to produce a term of type $\ttype{Nat}{\tunit}$ as the result of applying an operator not associated with the family $\fvar{Nat}$ is if that operator extracts the term from an argument (e.g.
+when projecting it out of an $n$-tuple), in which case the necessary invariants are inductively  maintained (see Sec.
+\ref{safety}).
+Translations extracted from denotations via \textsf{case} analysis are tracked during this phase as \emph{abstract internal terms}, $\tvalof{\tau_1}{\tau_2}$ (see rule \textsc{dencase-eval-valid}).%
+
+Let us first review how expressions are assigned types and abstract translations.
 Variables (\textsc{att-var}) translate directly to variables in the internal language, with the type determined by the typing context, $\Gamma$.
-Lambda terms (\textsc{att-lam}) also translate to lambda terms in the internal language and are given the appropriate $\fvar{Arrow}$ type.
-The internal type of the argument is $\trepof{\tau_1'}$, which is called the \emph{abstract internal type} associated with $\tau_1'$ (which is the argument type $\tau_1$ after it has been fully evaluated to a type-level value).
-This will be converted to the actual internal type determined by $\tau'$'s representation schema during deabstraction.
-Otherwise, typechecking and translation proceeds recursively be extending the typing context as is usual.
-
-The most interesting case is operator application (\textsc{att-op}).
-This rule extracts the appropriate operator definition, evaluates the operator index to a value, $\tauidx'$, and then recursively assigns a type and abstract translation to each argument.
-Next, it constructs a list of denotations from these and passes it, along with the fully evaluated operator index, to the implementation of the operator, $\tauop$.
-If this evaluates to a valid denotation, then compilation can proceed.
-Otherwise, compilation will stop, here because no other rule will apply, but in practice by displaying a type error for the user.
-
-The resulting valid denotation cannot be used directly, however.
-First, we must check that it is \emph{representationally consistent}, meaning that the abstract internal term produced, $\gabs$, can be shown to have the abstract internal type associated with the type, $\tau$, that was assigned to the expression.
-The judgement $\delfromtau{Fam}{\fvalCtx}{\tau}{\sabs}$ determines the abstract internal type, and $\checkRC{\gtCtx}{Fam}{\fvalCtx}{\gabs}{\sabs}$ performs the check.
-To do so, the variable context $\Gamma$, which associates variables with types, must be converted to a context associating variables with their corresponding abstract internal type, $\gtCtx$, as specified by the judgement $\eCtxTogCtx{Fam}{\fvalCtx}{\etCtx}{\gtCtx}$.
-Each of these judgements are defined in Fig.
+Lambda terms (\textsc{att-lam}) also translate to lambda terms in the internal language and are given the appropriate $\fvar{Arrow}$ type by extending the typing context and proceeding recursively as is usual.
+The internal type of the argument, however, is left as $\trepof{\tau_1'}$, which is called the \emph{abstract representation} of $\tau_1'$.
+The actual representation of this type is only available from within operators in its family.
+
+Now we will consider the important operator application rule (\textsc{att-op}).
+This rule operates by extracting the appropriate operator definition, evaluating the operator index to a value, $\tauidx'$, and recursively assigning a type and abstract translation to each argument.
+From these, it constructs a list of denotations  and passes this list, along with the fully evaluated operator index, to the  operator implementation, $\tauop$.
+If this results in a valid denotation, compilation can proceed, but if an $\terr$ is produced, compilation will stop because no other rule will apply (in practice, we would display a type error at this point).
+
+Before a valid denotation produced in this way can be used, however, we check that it is \emph{representationally consistent}, meaning that the abstract translation, $\gabs$, is of an abstract internal type consistent with the abstraction representation of the type it is paired with, $\tau$.
+The premise $\ddbar{\Xi_0,\fvar{Fam}}{\fvalCtx}{\trepof{\tau}}{\sabs}$ determines the abstract internal type and the premise $\checkRC{\gtCtx}{\Xi_0,\fvar{Fam}}{\fvalCtx}{\gabs}{\sabs}$ performs the internal type checking.
+To do so, the variable context, $\Gamma$, which associates variables with types, must be converted to a context associating those variables with their corresponding abstract internal types, $\gtCtx$.
+The relevant judgements are defined in Fig.
 \ref{ait}.
-
-To make this fundamental mechanism clear, and explain the rule \textsc{dencase-eval-valid} mentioned above, let us go through how the successor operator from Fig.
-\ref{example} works during this phase.
-
-The special internal term $\tvalof{\titerm{\gamma}}{\tautype}$ produced when the rule $\textsc{dencase-eval-valid}$ is applied is what allows the actual term produced by an operator outside of the local family to be held abstract.
+The schema context $\Xi$ is used to track which representation schemas are visible.
+In our calculus, $\Xi_0=\fvar{Arrow}$ is always visible alongside the schema of the family associated with the operator being considered.
+This could be exanded to support module-scoped visibility (we do not include modules in our calculus for simplicity), mutually-defined type families or type families that intentionally expose their representation schemas publicly (as $\fvar{Arrow}$ does).
+
+To clarify this fundamental mechanism let us examine how the expression $\eop{Nat}{s}{\tunit}{\eop{Nat}{z}{\tunit}{ }}$ will be processed during this phase.
+The inner application of $\opvar{z}$ will produce the abstract translation $0$ paired with the type $\ttype{Nat}{\tunit}$.
+Because the representation schema for $\fvar{Nat}$ is available, we have that $\ddbar{\Xi_0,\fvar{Nat}}{\fvalCtx}{\trepof{\ttype{Nat}{\tunit}}}{\dint}$ by \textsc{show-rep}, as needed.
+In contrast, if $\opvar{badnat}$ were used inside, the representation schema of $\fvar{Nat}$ would not be available, so we can only apply \textsc{hide-rep}, which does not give us enough information to give $-1$ a type.
+
+When the compiler next considers the outer application of $\opvar{s}$, it will pass the denotation $\tden{\titerm{0}}{\ttype{Nat}{\tunit}}$ into its definition.
+There, it binds the translation to the variable $\tvar{x}$ (within the $\tvar{pop\_final}$ helper function).
+However, $\tvar{x}$ is not simply $\titerm{0}$.
+Instead, its provenance is tracked by using the form $\tvalof{\titerm{0}}{\ttype{Nat}{\tunit}}$.
+When attempting to check that $\iup{\tvar{x}} + 1$ is valid, the \textsc{show-trans} rule will reveal that it is an integer because, again, the appropriate representation schema is available.
+If the schema were not available, the most that could be derived is that $\checkRC{\gtCtx}{\Xi}{\fvalCtx}{\iup{\tvar{x}}}{\trepof{\ttype{Nat}{\tunit}}}$.
+The fact that natural numbers are represented using integers is not derivable (though it is true), so the addition operation would fail to type.
+This fact would, however, be sufficient for implementing families, like $\fvar{Ntuple}$, that do not require knowledge about a type's representation.
+We encourage the reader to derive these judgements for the logic in the $\opvar{rec}$ operator to strengthen their understanding of this fundamental mechanism.
+
+We cannot leave abstract representations and abstract internal terms in the result of compilation, so a final deabstraction phase erases these, replacing them with their underlying internal types and terms in all contexts.
+Just as with abstract types in modules or existential types, there is no run-time overhead to this mechanism -- programs run at full speed.
+We give the deabstraction rules in the appendix due to their simplicity.
 
@@
 
-\begin{figure}
+\begin{figure}[t]
 \small
+$\fbox{\inferrule{}{\concrep{\fvalCtx}{\tau}{\sigma}}}$
+\begin{mathpar}
+\inferrule[get-rep]{
+	\fval{Fam}{\theta}{i}{\tau} \in \fvalCtx\\
+	\tEvalX{[\tauidx/\tvar{i}]\tau}{\titype{\sigma}}\\
+	\ddbarX{\sigma}{\sconc}
+}{
+	\concrep{\fvalCtx}{\ttype{Fam}{\tauidx}}{\sconc}
+}
+\end{mathpar}
 $\fbox{\inferrule{}{\ddbarX{\sigma}{\sigma'}}}$
-~~~$\fvalCtx ::= \fvalCtx_0 \pipe \fvalCtx, \fvalDf$
-\begin{mathpar}
-\inferrule[abs int]{ }{
+~~~$\fvalCtx ::= \fvalCtx_0 \pipe \fvalCtx, \fvalDf$~~~
+~~~$\Xi ::= \Xi_0 \pipe \Xi, \fvar{Fam}$~~~
+~~~$\Xi_0 := \fvar{Arrow}$
+\begin{mathpar}
+\inferrule[abs-int]{ }{
 	\ddbarX{\dint}{\dint}
@@
 
-\inferrule[abs arrow]{
+\inferrule[abs-arrow]{
 	\ddbarX{\sigma_1}{\sigma_1'}\\
@@
 
-\inferrule[abs prod]{
+\inferrule[abs-prod]{
 	\ddbarX{\sigma_1}{\sigma_1'}\\
@@
 
-\inferrule[abs+cancel unquote]{
-	\ddbarX{\sigma}{\sigma'}
-}{
-	\ddbarX{\dup{\titype{\sigma}}}{\sigma'}
-}
-
-\inferrule[abs rep from type visible]{
-	\fvalDf \in \fvalCtx\\
-	\tEvalX{[\tauidx/\tvar{i}]\tau}{\titype{\sigma}}\\
-	\ddbarX{\sigma}{\sigma}
-}{
-	\ddbarX{\trepof{\ttype{Fam}{\tauidx}}}{\sigma}
-}
-
-\inferrule[abs rep from type hidden]{
-	\fvar{Fam} \neq \fvar{Fam'}
-}{
-	\ddbarX{\trepof{\ttype{Fam'}{\tauidx}}}{\trepof{\ttype{Fam'}{\tauidx}}}
-}
-\end{mathpar}
-$\fbox{\inferrule{}{\delfromtau{Fam}{\fvalCtx}{\tau}{\sigma}}}$
-\begin{mathpar}
-\inferrule[abs rep from type]{
-	\fval{Fam'}{\theta}{i}{\tau} \in \fvalCtx\\
-	\tEvalX{[\tauidx/\tvar{i}]\tau}{\titype{\sigma}}\\
-	\ddbarX{\sigma}{\sigma}
-}{
-	\tauisdel{Fam}{\fvalCtx}{Fam'}{\tauidx}{\sigma}
+\inferrule[itype-inverse]{
+	\ddbarX{\sigma}{\sabs}
+}{
+	\ddbarX{\dup{\titype{\sigma}}}{\sabs}
+}
+
+\inferrule[show-rep]{
+	\fvar{Fam} \in \Xi\\
+	\concrep{\fvalCtx}{\ttype{Fam}{\tauidx}}{\sconc}
+}{
+	\ddbarX{\trepof{\ttype{Fam}{\tauidx}}}{\sconc}
+}
+
+\inferrule[hide-rep]{
+	\fvar{Fam} \notin \Xi
+}{
+	\ddbarX{\trepof{\ttype{Fam}{\tauidx}}}{\trepof{\ttype{Fam}{\tauidx}}}
 }
@@
 \begin{mathpar}
-\inferrule[abs empty]{ }{
+\inferrule[abs-empty]{ }{
 	\eCtxTogCtxX{\emptyctx}{\emptyctx}
@@
 
-\inferrule[abs ctx]{
+\inferrule[abs-ctx]{
 	\eCtxTogCtxX{\etCtx}{\gtCtx}\\
-	\delfromtau{Fam}{\fvalCtx}{\tau}{\sigma}
+	\ddbarX{\trepof{\tau}}{\sigma}
 }{
@@
 \begin{mathpar}
-\inferrule[abs i-var]{ }{
-	\checkRC{\gtCtxX{x}{\sigma}}{Fam}{\fvalCtx}{\evar{x}}{\sigma}
-}
-
-\inferrule[abs i-lam]{
-	\ddbarX{\sigma_1}{\sigma_1}\\
-	\checkRC{\gtCtxX{x}{\sigma_1}}{Fam}{\fvalCtx}{\gamma}{\sigma_2}
-}{
-	\checkRCX{\ilam{x}{\sigma_1}{\gamma}}{\darrow{\sigma_1}{\sigma_2}}
-}
-
-\inferrule[abs i-ap]{
+\inferrule[abs-i-var]{ }{
+	\checkRC{\gtCtxX{x}{\sigma}}{\Xi}{\fvalCtx}{\evar{x}}{\sigma}
+}
+
+\inferrule[abs-i-lam]{
+	\ddbarX{\sigma_1}{\sigma_1'}\\
+	\checkRC{\gtCtxX{x}{\sigma_1'}}{\Xi}{\fvalCtx}{\gamma}{\sigma_2}
+}{
+	\checkRCX{\ilam{x}{\sigma_1}{\gamma}}{\darrow{\sigma_1'}{\sigma_2}}
+}
+
+\inferrule[abs-i-ap]{
 	\checkRCX{\gamma_1}{\darrow{\sigma_1}{\sigma_2}}\\
@@
 
-\inferrule[abs i-fix]{
-	\ddbarX{\sigma}{\sigma}\\
-	\checkRC{\gtCtxX{x}{\sigma}}{Fam}{\fvalCtx}{\gamma}{\sigma}
-}{
-	\checkRCX{\ifix{x}{\sigma}{\gamma}}{\sigma}
+\inferrule[abs-i-fix]{
+	\ddbarX{\sigma}{\sigma'}\\
+	\checkRC{\gtCtxX{x}{\sigma'}}{\Xi}{\fvalCtx}{\gamma}{\sigma'}
+}{
+	\checkRCX{\ifix{x}{\sigma}{\gabs}}{\sigma'}
 }
@@
 %
-\inferrule[abs if eq]{
-	\checkRCX{\gamma_1}{\dint}\\
-	\checkRCX{\gamma_2}{\dint}\\\\
-	\checkRCX{\gamma_3}{\sigma}\\
-	\checkRCX{\gamma_4}{\sigma}
-}{
-	\checkRCX{\iIfEq{\gamma_1}{\gamma_2}{\dint}{\gamma_3}{\gamma_4}}{\sigma}
-}
-
-\inferrule[abs iterm unquote]{
+% \inferrule[abs-int-eq]{
+% 	\checkRCX{\gamma_1}{\dint}\\
+% 	\checkRCX{\gamma_2}{\dint}\\\\
+% 	\checkRCX{\gamma_3}{\sigma}\\
+% 	\checkRCX{\gamma_4}{\sigma}
+% }{
+% 	\checkRCX{\iIfEq{\gamma_1}{\gamma_2}{\dint}{\gamma_3}{\gamma_4}}{\sigma}
+% }
+\inferrule[abs-iterm-inverse]{
 	\checkRCX{\gamma}{\sigma}
@@
 
-\inferrule[abs val from den visible]{
-	\tauisdel{Fam}{\fvalCtx}{Fam}{\tauidx}{\sigma}\\
+\inferrule[show-trans]{
+	\fvar{Fam} \in \Xi\\
+	\concrep{\fvalCtx}{\ttype{Fam}{\tauidx}}{\sigma}\\
 	\checkRCX{\gamma}{\sigma}
@@
 
-\inferrule[abs val from den hidden]{
-	\fvar{Fam} \neq \fvar{Fam'}\\
-	\tauisdel{Fam}{\fvalCtx}{Fam'}{\tauidx}{\sigma}\\
+\inferrule[hide-trans]{
+	\fvar{Fam} \notin \Xi\\
+	\concrep{\fvalCtx}{\ttype{Fam}{\tauidx}}{\sigma}\\
 	\checkRCX{\gamma}{\sigma}
 }{
-	\checkRCX{\tvalof{\titerm{\gamma}}{\ttype{Fam'}{\tauidx}}}{\trepof{\ttype{Fam'}{\tauidx}}}
+	\checkRCX{\tvalof{\titerm{\gamma}}{\ttype{Fam}{\tauidx}}}{\trepof{\ttype{Fam}{\tauidx}}}
 }
@@
 \label{ait}
-\end{figure}
-\begin{figure}[t]
-\small
-$\fbox{\inferrule{}{\eraseX{\sigma}{\sigma}}}$
-\begin{mathpar}
-\inferrule[deabs int]{ }{
-	\eraseX{\dint}{\dint}
-}
-
-\inferrule[deabs arrow]{
-	\eraseX{\sigma_1}{\sigma_1}\\\\
-	\eraseX{\sigma_2}{\sigma_2}
-}{
-	\eraseX{\darrow{\sigma_1}{\sigma_2}}{\darrow{\sigma_1}{\sigma_2}}
-}
-
-\inferrule[deabs prod]{
-	\eraseX{\sigma_1}{\sigma_1}\\\\
-	\eraseX{\sigma_2}{\sigma_2}
-}{
-	\eraseX{\dpair{\sigma_1}{\sigma_2}}{\dpair{\sigma_1}{\sigma_2}}
-}
-
-\inferrule[deabs  rep from type hidden]{
-	\fvalDf \in \fvalCtx\\
-	\tEvalX{[\tauidx/\tvar{i}]\tau}{\titype{\sigma}}\\\\
-	\ddbarX{\sigma}{\sigma}\\
-	\eraseX{\sigma}{\sigma}
-}{
-	\eraseX{\trepof{\ttype{Fam}{\tauidx}}}{\sigma}
-}
-\end{mathpar}
-$\fbox{\inferrule{}{\eraseX{\gamma}{\gamma}}}$
-\begin{mathpar}
-\inferrule[deabs i-var]{ }{
-	\eraseX{\evar{x}}{\evar{x}}
-}
-
-\inferrule[deabs i-lam]{
-	\ddbar{\_}{\fvalCtx}{\sigma}{\sigma}\\\\
-	\eraseX{\sigma}{\sigma}\\\\
-	\eraseX{\gamma}{\gamma}
-}{
-	\eraseX{\ilam{x}{\sigma}{\gamma}}{\ilam{x}{\sigma}{\gamma}}
-}
-
-%\inferrule[deabs ap]{
-%	\eraseX{\gamma_1}{\gamma_1}\\
-%	\eraseX{\gamma_2}{\gamma_2}
-%}{
-%	\eraseX{\iapp{\gamma_1}{\gamma_2}}{\iapp{\gamma_1}{\gamma_2}}
-%}
-%
-\inferrule[deabs i-fix]{
-	\ddbar{\_}{\fvalCtx}{\sigma}{\sigma}\\\\
-	\eraseX{\sigma}{\sigma}\\\\
-	\eraseX{\gamma}{\gamma}
-}{
-	\eraseX{\ifix{x}{\sigma}{\gamma}}{\ifix{x}{\sigma}{\gamma}}
-}
-\\
-\text{\color{gray} (omitted forms have trivially recursive rules)}
-\\
-\inferrule[deabs unquote]{
-	\eraseX{\gamma}{\gamma}
-}{
-	\eraseX{\iup{\titerm{\gamma}}}{\gamma}
-}
-
-\inferrule[deabs val from den hidden]{
-	\eraseX{\gamma}{\gamma}
-}{
-	\eraseX{\tvalof{\titerm{\gamma}}{\ttype{Fam}{\tauidx}}}{\gamma}
-}
-\end{mathpar}
-\caption{\small Deabstraction rules}
-\label{deabs}
 \end{figure}
@@
 \section{Safety of @$\lambda$}\label{safety}
+In giving users of a language direct influence on the typechecking and translation of expressions, it is essential to consider safety properties.
+It must not be possible for well-typed programs to fail to finish compiling or go wrong at run-time because of a buggy extension.
+It is also considered desirable for typechecking to be decidable.
+And for extensions to be reliable, they must not be allowed to interfere with one another under any circumstance.
+By carefully designing our extension mechanism, we believe we have achieved each of these goals.
+
+\subsection{Representational Consistency and Type Safety}
+We do not need to give a semantics to internal terms and internal types that do not survive deabstraction.
+The remaining terms and types form a variant of PCF with well-understood primitives (here, integers and binary products) known to be type safe \cite{pfpl}.
+If compilation can be shown to always result in a well-typed term of this language, then type safety follows by composing these two facts.
+Fortunately, this fact is a corollary of our representational consistency mechanism, described in Sec.
+\ref{repcon}.
+The proof is by straightforward (though deeply nested!) rule induction.
+
+\begin{theorem}[Representational Consistency]
+If $\progOK{ }{\Sigma}{e}$ and $\ecompiles{ }{\fvalCtx}{e}{\tau}{\gabs}$ and $\eraseX{\gabs}{\gamma}$ and 	$\ddbar{\Xi_0}{\fvalCtx}{\trepof{\tau}}{\sabs}$ and $\eraseX{\sabs}{\sigma}$ then $\checkRC{ }{\Xi_0}{\fvalCtx}{\gamma}{\sigma}$.
+\end{theorem}
+
+\subsection{Decidability}
+To show that compilation is decidable, we need to show that both kind checking and active typechecking and compilation are decidable.
+This hinges on showing that the type-level language is typesafe and that evaluation of type-level terms always terminates.
+
+\begin{theorem}[Kind Safety and Termination]
+If $\vdash_{\Sigma} \tau : \kappa$ then $\tEvalX{\tau}{\tau'}$ such that $\vdash_{\Sigma_0} \tau' : \kappa$ and $\tEvalX{\tau'}{\tau'}$. 
+\end{theorem}
+
+The proof is straightforward because the type-level language is based on a conventional simply-typed lambda calculus with only bounded recursion over lists, so standard techniques (e.g.
+logical relations) can be used directly.
+The only new constructs are the types, denotations and reified internal language forms, but because none of these can contain type-level functions by the statics, there is no risk of introducing self-reference by their inclusion.
+If the type-level language included constructs like inductive datatypes (without a strict positivity condition) or general recursion, this theorem would be weakened.
+
+\subsection{Non-Interference}
+Our abstraction mechanism guarantees that the representation invariants collectively maintained by the operators associated with a type family cannot be violated by operators in other type families, by ensuring that introductory forms cannot be defined outside the family.
+One way to state this is as follows:
+
+\begin{theorem}[Non-Interference]
+For any property $P(\gamma)$, if $\progOK{ }{\Sigma}{e}$ and $\ecompiles{ }{\fvalCtx}{e}{\tau}{\gabs}$ and $\eraseX{\gabs}{\gamma}$ implies $P(\gamma)$, then for any $\Sigma'$ and $\fvalCtx'$ disjoint from $\Sigma$ and $\fvalCtx$, we have that if $\progOK{ }{\Sigma,\Sigma'}{e'}$ and $\ecompiles{ }{\fvalCtx,\fvalCtx'}{e'}{\tau}{\gabs'}$ and $\eraseX{\gabs'}{\gamma'}$ then $P(\gamma')$.
+\end{theorem}
+The proof relies on the fact that the show/hide rules are mutually exclusive and so there are two cases to consider for each form of expression, and weakening properties of the family contexts.
+This powerful theorem allows us to compose type families arbitrarily without needing to handle cases in our implementation that are ruled out based on a local analysis.
+The details will be provided in the appendix.
+
+% \subsection{Modular Verification}
+% nope
 
@@
 
-\section{Examples}\label{examples}
-
-\section{Design Considerations}\label{design}
+\section{Use Cases}\label{examples}
+Indexed type families and indexed operator families are a powerful tool for programming language semantics.
+In \emph{Practical Foundations for Programming Languages}, for example, nearly every chapter defines a new collection of indexed types and operators and gives their semantics in isolation from the constructs in the other chapters \cite{pfpl}.
+For example, $n$-ary sums and products are families indexed by a list of types.
+Labeled variants of these simply pair the labels with the types as static strings.
+Even nested pattern matching, such as that found in modern functional languages, can be understood as an operator family indexed by a series of patterns, which can be represented as type-level data.
+In the appendix, we show how to implement sums and products in \atlam.
+
+The book also shows constructs as varied as dynamic types as well as a number of constructs for parallelism and concurrency, such as futures, promises and actors.
+With a sufficiently capable internal language (exposing basic concurrency primitives, for example), each of these could be fully implemented as libraries using our mechanism.
+Few languages include primitive support for several such abstractions, even though they are often useful together.
+In a preliminary implementation of this calculus, Ace, which is beyond the scope of this paper, we have implemented the entirety of the OpenCL programming language as a library, along with primitives supporting partitioned global address spaces.
+
+Object systems too could be implemented using this mechanism, with indices serving to capture the inheritance data and the signatures.
+Operator families for reading and writing fields and/or sending messages would be parameterized by strings naming them.
+The operator definition would simply search the signatures going up the inheritance hierarchy, and implement objects in a conventional way (e.g.
+using a v-table together with a pointer to a structure containing field data).
+A variety of object systems could coexist within the same language.
+Another related example would be to implement a safe and efficient interoperability layer with an existing OO language by capturing its type system as an extension, to be used only at language boundaries.
+
+Finally, a variety of domain-specific type systems, capturing complex rule systems like the system of scientific units of measure (a family indexed by the measure being used and the type which the measure is being applied to), regular expressions \cite{regexp} (a family indexed by the number of captured groups) or XML \cite{xml} (indexed by a document schema) would be possible.
+Recent work examined a specialized type system capturing the semantics of the widely-used jQuery javascript library \cite{jquery-oopsla}.
+Today, these all require \emph{ad hoc} language-external solutions, or the development of new languages.
+
+%\section{Design Considerations}\label{design}
 
 \section{Related Work}\label{related-work}
-Our representation schema abstraction mechanism is purely syntactic, so it relates to previous work on syntactic type abstraction \cite{syntypeabs}.
-While that work was focused on abstracting away the identity of a particular type outside of the ``principal'' it is associated with, we focus on abstracting away the knowledge of how a primitive type family is implemented outside of its operators.
-
-Compiler correctness / ML work (Harper, Crary).
-
-Erdweg's work (see email)
-
-Term rewriting systems
-
-\subsection{Type-Level Computation} %Haskell, Ur and $\Omega$mega
-System XX with simple case analysis provides the basis of type-level computation in Haskell (where type-level functions are called type families \cite{Chakravarty:2005:ATC}).
+Our representation schema abstraction mechanism relates closely to abstract and existential types \cite{pfpl,atpl}.
+Our calculus enforces the abstraction barriers  in a purely syntactic manner, as in previous work on syntactic type abstraction \cite{syntypeabs}.
+While this work is all focused on abstracting away the identity of a particular type outside of the ``principal'' it is associated with (e.g.
+a module), we focus on abstracting away the knowledge of how a primitive type family is implemented outside of a limited scope.
+
+The representational consistency mechanism brings into the language work on typed compilation, especially work done on Standard ML in the TILT project \cite{tilt}.
+Indeed, the specification of Standard ML is structured around a typed internal language and a judgement that assigns a type and an internal term to each expression \cite{smlstd}.
+Representational consistency is related to the notion of type-directed copmilation in this work.
+
+Type-level computation is supported in some form by a growing number of languages.
+For example, Haskell supports a simple form of it \cite{Chakravarty:2005:ATC}).
 Ur uses type-level records and names to support typesafe metaprogramming, with applications to web programming \cite{conf/pldi/Chlipala10}.
@@
 Dependently-typed languages blur the traditional phase separation between types and expressions, so type-level computation is often implicitly used (though not always in its most general form, e.g.
-Deputy \cite{conf/icfp/ChenX05}, ATS \cite{conf/esop/ConditHAGN07}.)
+Deputy \cite{conf/icfp/ChenX05}, ATS \cite{conf/esop/ConditHAGN07}).
+We show how to integrate language extensions into the type-level language, drawing on ideas about \cite{activelibraries}.
 
@@
 
-\subsection{Term Rewriting Systems}
 Many languages and tools allow developers to rewrite expressions according to custom rules.
@@
 MetaML \cite{Sheard:1999:UMS}), and external rewrite systems also exist for many languages.
-These facilities differ from type-level specification in one or more of the following ways:
-
-\begin{enumerate}
-\item In type-level specification, the type of a value is determined separately from its representation; in fact, the same representation may be generated by multiple types. 
-\item We draw a distinction between the metalanguage, used to specify types and compile-time logic, the source grammar, used to describe run-time behavior, and the internal language, used to implement this behavior.
-Term rewriting systems generally do not draw this distinction.
-By doing so, each component language can be structured and constrained as appropriate for its distinct role, as we show.
+These facilities differ from AT\&T in that they involve direct manipulation of terms, while AT\&T involves extending typechecking and translation logic directly.
+We also draw a distinction between the type-level, used to specify types and compile-time logic, the expression grammar, used to describe run-time behavior, and the internal language, used to implement this behavior.
+By doing so, each component can be structured and constrained as appropriate for its distinct role, as we have shown.
+
+Previous work on extensible languages has suffered from problems with either expressiveness or safety.
+For example, a number of projects, such as SugarJ \cite{erdweg2011sugarj}, allow for user-defined desugarings (and indeed, our system would clearly benefit from integration with such a mechanism), but this does not allow the semantics to be fundamentally extended nor for implementation details to be hidden.
+Recent variants of this work has investigated introducing new typing rules, but only if they are admissible by the base type system \cite{erdwegicfp}.
+Our work allows for entirely new logic to be added to the system, requiring only that the implementation of this logic respect the internal type system.
+A number of other extensible languages (e.g.
+Xroma \cite{xroma}) and compilers do allow new static semantics to be introduced, but in an unconstrained manner based on global pattern matching and rewriting, leading to significant problems with interference and safety.
+Our work aims to provide a sound and safe underpinning, based around well-understood concepts of type and operator families, to language extensibility.
+
+In the future, we will investigate mechanisms to enable type systems with specialized binding and scoping rules, as well as integration of dependent kinds to support mechanized verification of key properties like representational consistency and adequacy against a declarative specification at kind-checking time.
+We will also investigate mechanisms that enable a more natural syntax (e.g.
+Wyvern's type-directed syntax \cite{globaldsl}).
+
 %\item With type-level specification, dispatch to a type-level function occurs implicitly on the basis of the structure of an expression.
@@
 A*\cite{fowler2010domain}, EPP\cite{fowler2010domain}) and macro systems for object-oriented languages, like OpenC++ \cite{fowler2010domain} and OpenJava \cite{fowler2010domain}, do offer a somewhat limited form of operation-based dispatch.
-\item Many common macro systems and metaprogramming facilities operate at run-time.
-Compilers for some forms of LISP employ aggressive compile-time specialization techniques to attempt to minimize this overhead.
-Static and staged term-rewriting systems also exist (e.g.
-OpenJava\cite{TatM:OpenJCBMSJ}, Template Haskell\cite{SheardPeytonJones:Haskell-02}, MetaML \cite{Sheard:1999:UMS} and others).
-\end{enumerate}
-
-\subsection{Language Frameworks}
-When the mechanisms available in an existing language prove insufficient, researchers and domain experts must design a new language.
+
+% \subsection{Language Frameworks}
+% When the mechanisms available in an existing language prove insufficient, researchers and domain experts must design a new language.
 A number of tools have been developed to assist with this task, including compiler generators, language workbenches and domain-specific language frameworks (cf \cite{fowler2010domain}).
 
-A major barrier to adoption is the fact that interoperability is intrinsically problematic.
+% A major barrier to adoption is the fact that interoperability is intrinsically problematic.
 Even languages which target a common platform, such as the Java Virtual Machine, can only interact using its limited set of primitives.
@@
 
-Instead of focusing on defining standalone languages, type-level specification gives greater responsibility in a granular manner to libraries.
+% Instead of focusing on defining standalone languages, type-level specification gives greater responsibility in a granular manner to libraries.
 In this way, a range of constructs can coexist within the same program and, assuming that it can be shown by some method that various constructs are safely composable, be mixed and matched.
@@
 Nevertheless, as Ace shows, a simple grammar can be used quite flexibly.
-\subsection{Extensible Compilers}
-An alternative methodology is to implement language features granularly as compiler extensions.
+% \subsection{Extensible Compilers}
+% An alternative methodology is to implement language features granularly as compiler extensions.
 As discussed in Section 1, existing designs suffer from the same problems related to composability, modularity\-, safety and security as extensible languages, while also adding the issue of language fragmentation.
 
-Type-level specification can in fact be implemented within a compiler, rather than provided as a core language feature.
+% Type-level specification can in fact be implemented within a compiler, rather than provided as a core language feature.
 This would resolve some of the issues, as described in this paper.
@@
 We also avoid the fragmentation issue.
-\subsection{Specification Languages}
-Several {\it specification languages} (or {\it logical frameworks}) based on these theoretical formulations exist, including the OBJ family of languages (e.g.
+% \subsection{Specification Languages}
+% Several {\it specification languages} (or {\it logical frameworks}) based on these theoretical formulations exist, including the OBJ family of languages (e.g.
 CafeOBJ \cite{Diaconescu-Futatsugi01}).
@@
 
-Type-determined compilation takes a more concrete approach to the problem, focusing on combining {\it implementations} of different\- logics, rather than simply their specifications.
+% Type-determined compilation takes a more concrete approach to the problem, focusing on combining {\it implementations} of different\- logics, rather than simply their specifications.
 In other words, it focuses on combining {\it type checkers} and {\it implementation strategies} rather than more abstract representations of a language's type system and dynamic semantics.
@@
 
-CANT GUARANTEE THAT SPECIFICATIONS ARE ACTUALLY DECIDABLE 
-\section{Discussion}\label{discussion}
+% CANT GUARANTEE THAT SPECIFICATIONS ARE ACTUALLY DECIDABLE 
 \bibliographystyle{abbrv}
@@
 \section{Appendix}
-\subsection{Statics of G\"odel's T}\label{T}
-\newcommand{\lam}[3]{\lambda #1{:}#2.#3}
-\newcommand{\ap}[2]{#1~#2}
-\newcommand{\z}{\textsf{z}}
-\newcommand{\s}[1]{\textsf{s}(#1)}
-\newcommand{\natrec}[5]{\textsf{natrec}(#1; #2; #3,#4.#5)}
-\newcommand{\pair}[2]{(#1, #2)}
-\newcommand{\fst}[1]{\textsf{fst}(#1)}
-\newcommand{\snd}[1]{\textsf{snd}(#1)}
-
-\newcommand{\tArrow}[2]{#1 \rightarrow #2}
-\newcommand{\nat}{\texttt{nat}}
-\renewcommand{\prod}[2]{#1 \times #2}
-
-%\newcommand{\eCtx}{\Gamma}
-\newcommand{\eCtxX}[3]{#1, #2 : #3}
-\newcommand{\jet}[3]{#1 \vdash #2 : #3}
-\newcommand{\jetX}[2]{\jet{\eCtx}{#1}{#2}}
-%\begin{figure}
-%\small
-\begin{mathpar}
-\small
-\inferrule[var]{ }{
-	\jet{\eCtxX{\eCtx}{x}{\tau}}{x}{\tau}
-}
-
-\inferrule[Arrow-I]{
-	\jet{\eCtxX{\eCtx}{x}{\tau}}{e}{\tau'}	
-}{
-	\jetX{\lam{x}{\tau}{e}}{\tArrow{\tau}{\tau'}}
-}
-
-\inferrule[Arrow-E]{
-	\jetX{e_1}{\tArrow{\tau}{\tau'}}\\
-	\jetX{e_2}{\tau}
-}{
-	\jetX{\ap{e_1}{e_2}}{\tau'}
-}
-
-\inferrule[Nat-I1]{ }{
-	\jetX{\z}{\nat}
-}
-~~~~~~~~
-\inferrule[Nat-I2]{
-	\jetX{e}{\nat}
-}{
-	\jetX{\s{e}}{\nat}
-}
-~~~~~~~~
-\inferrule[Nat-E]{
-	\jetX{e_1}{\nat}\\
-	\jetX{e_2}{\tau}\\
-	\jetX{e_3}{\tArrow{\nat}{\tArrow{\tau}{\tau}}}
-}{
-	\jetX{\natrec{e_1}{e_2}{x}{y}{e_3}}{\tau}
-}
-\end{mathpar}
-%\caption{Static semantics of G\"odel's T}
-%\end{figure}
-% \begin{figure}
-% \small
-% \begin{mathpar}
-% \inferrule[Prod-I]{
-% 	\jetX{e_1}{\tau_1}\\
-% 	\jetX{e_2}{\tau_2}
-% }{
-% 	\jetX{\pair{e_1}{e_2}}{\prod{\tau_1}{\tau_2}}
-% }
-
-% \inferrule[Prod-E1]{
-% 	\jetX{e}{\prod{\tau_1}{\tau_2}}
-% }{
-% 	\jetX{\fst{e}}{\tau_1}
-% }
-
-% \inferrule[Prod-E2]{
-% 	\jetX{e}{\prod{\tau_1}{\tau_2}}
-% }{
-% 	\jetX{\snd{e}}{\tau_2}
-% }
-% \end{mathpar}
-% \caption{Static semantics of products}\label{prods}
-% \end{figure}
-\subsection{Helper Functions}\label{helper}
-\todo{describe these}
-\[\begin{array}{rcl}
-\tvar{const} & := & (\tlam{a}{\klist{\kDen}}{\tlam{d}{\kDen}{\tfold{\tvar{a}}{\tvar{d}}{\_}{\_}{\_}{\terr}}})\\
-\tvar{pop} & := & (\tlam{a}{\klist{\kDen}}{\tlam{f}{\karrow{\kITerm}{\karrow{\kTypeBlur}{\karrow{\klist{\kDen}}{\kDen}}}}{\\& & \quad\tfold{\tvar{a}}{\terr}{d}{b}{\_}{\tdencase{\tvar{d}}{x}{t}{
-	\tapp{\tapp{\tapp{\tvar{f}}{\tvar{x}}}{\tvar{t}}}{\tvar{b}}}{\terr}}}})\\
-\tvar{pop\_final} & := & (\tlam{a}{\klist{\kDen}}{\tlam{f}{\karrow{\kITerm}{\karrow{\kTypeBlur}{\kDen}}}{\\& & \quad\tfold{\tvar{a}}{\terr}{d}{b}{\_}{\\& & \quad \tfold{\tvar{b}}{\tdencase{\tvar{d}}{x}{t}{\tapp{\tapp{\tvar{f}}{\tvar{x}}}{\tvar{t}}}{\terr}}{\_}{\_}{\_}{\terr}}}})\\
-\tvar{check\_type} & := & (\tlam{t1}{\kTypeBlur}{\tlam{t2}{\kTypeBlur}{\tlam{d}{\kDen}{\tifeq{\tvar{t1}}{\tvar{t2}}{\kTypeBlur}{\tvar{d}}{\terr}}}})
-\end{array}\]
+\verb|http://www.cs.cmu.edu/~comar/att-esop14-appendix.pdf|.
 \end{document}

### 2013-11-07 ace-pldi14/ace-pldi14.tex
Initial port of PLDI14 paper from SC13 folder

--- 2013-04-28 ace/ace.tex
@@
-%% bare_conf.tex
-%% V1.3
-%% 2007/01/11
-%% by Michael Shell
-%% See:
-%% http://www.michaelshell.org/
-%% for current contact information.
-%%
-%% This is a skeleton file demonstrating the use of IEEEtran.cls
-%% (requires IEEEtran.cls version 1.7 or later) with an IEEE conference paper.
-%%
-%% Support sites:
-%% http://www.michaelshell.org/tex/ieeetran/
-%% http://www.ctan.org/tex-archive/macros/latex/contrib/IEEEtran/
-%% and
-%% http://www.ieee.org/
-
-%%*************************************************************************
-%% Legal Notice:
-%% This code is offered as-is without any warranty either expressed or
-%% implied; without even the implied warranty of MERCHANTABILITY or
-%% FITNESS FOR A PARTICULAR PURPOSE! 
-%% User assumes all risk.
-%% In no event shall IEEE or any contributor to this code be liable for
-%% any damages or losses, including, but not limited to, incidental,
-%% consequential, or any other damages, resulting from the use or misuse
-%% of any information contained here.
-%%
-%% All comments are the opinions of their respective authors and are not
-%% necessarily endorsed by the IEEE.
-%%
-%% This work is distributed under the LaTeX Project Public License (LPPL)
-%% ( http://www.latex-project.org/ ) version 1.3, and may be freely used,
-%% distributed and modified. A copy of the LPPL, version 1.3, is included
-%% in the base LaTeX documentation of all distributions of LaTeX released
-%% 2003/12/01 or later.
-%% Retain all contribution notices and credits.
-%% ** Modified files should be clearly indicated as such, including  **
-%% ** renaming them and changing author support contact information. **
-%%
-%% File list of work: IEEEtran.cls, IEEEtran_HOWTO.pdf, bare_adv.tex,
-%%                    bare_conf.tex, bare_jrnl.tex, bare_jrnl_compsoc.tex
-%%*************************************************************************
-\documentclass{sig-alternate}
-
-%\usepackage{cite}
-%\renewcommand{\citepunct}{,\,} % IEEEtran wants to use ],\,[ for this but that looks dumb...
+%\documentclass[12pt]{article}
+\documentclass[9pt,preprint]{sigplanconf}
+% The following \documentclass options may be useful:
+%
+% 10pt          To set in 10-point type instead of 9-point.
+% 11pt          To set in 11-point type instead of 9-point.
+% authoryear    To obtain author/year citation style instead of numeric.
+\usepackage{listings}
+\usepackage[usenames,dvipsnames]{color}
+\usepackage{mathpartir}
+\usepackage{amsmath}
+\usepackage{amssymb} 
+\usepackage{amsthm}
+\usepackage{ stmaryrd }
 
@@
 
-% *** GRAPHICS RELATED PACKAGES ***
-%
-% *** MATH PACKAGES ***
-%
-%\usepackage[cmex10]{amsmath}
-% A popular package from the American Mathematical Society that provides
-% many useful and powerful commands for dealing with mathematics. If using
-% it, be sure to load this package with the cmex10 option to ensure that
-% only type 1 fonts will utilized at all point sizes. Without this option,
-% it is possible that some math symbols, particularly those within
-% footnotes, will be rendered in bitmap form which will result in a
-% document that can not be IEEE Xplore compliant!
-%
-% Also, note that the amsmath package sets \interdisplaylinepenalty to 10000
-% thus preventing page breaks from occurring within multiline equations. Use:
-%\interdisplaylinepenalty=2500
-% after loading amsmath to restore such page breaks as IEEEtran.cls normally
-% does. amsmath.sty is already installed on most LaTeX systems. The latest
-% version and documentation can be obtained at:
-% http://www.ctan.org/tex-archive/macros/latex/required/amslatex/math/
-
-% *** SPECIALIZED LIST PACKAGES ***
-%
-%\usepackage{algorithmic}
-% algorithmic.sty was written by Peter Williams and Rogerio Brito.
-% This package provides an algorithmic environment fo describing algorithms.
-% You can use the algorithmic environment in-text or within a figure
-% environment to provide for a floating algorithm. Do NOT use the algorithm
-% floating environment provided by algorithm.sty (by the same authors) or
-% algorithm2e.sty (by Christophe Fiorio) as IEEE does not use dedicated
-% algorithm float types and packages that provide these will not provide
-% correct IEEE style captions. The latest version and documentation of
-% algorithmic.sty can be obtained at:
-% http://www.ctan.org/tex-archive/macros/latex/contrib/algorithms/
-% There is also a support site at:
-% http://algorithms.berlios.de/index.html
-% Also of interest may be the (relatively newer and more customizable)
-% algorithmicx.sty package by Szasz Janos:
-% http://www.ctan.org/tex-archive/macros/latex/contrib/algorithmicx/
-
-% *** ALIGNMENT PACKAGES ***
-%
-%\usepackage{array}
-% Frank Mittelbach's and David Carlisle's array.sty patches and improves
-% the standard LaTeX2e array and tabular environments to provide better
-% appearance and additional user controls. As the default LaTeX2e table
-% generation code is lacking to the point of almost being broken with
-% respect to the quality of the end results, all users are strongly
-% advised to use an enhanced (at the very least that provided by array.sty)
-% set of table tools. array.sty is already installed on most systems. The
-% latest version and documentation can be obtained at:
-% http://www.ctan.org/tex-archive/macros/latex/required/tools/
-
-%\usepackage{mdwmath}
-%\usepackage{mdwtab}
-% Also highly recommended is Mark Wooding's extremely powerful MDW tools,
-% especially mdwmath.sty and mdwtab.sty which are used to format equations
-% and tables, respectively. The MDWtools set is already installed on most
-% LaTeX systems. The lastest version and documentation is available at:
-% http://www.ctan.org/tex-archive/macros/latex/contrib/mdwtools/
-
-% IEEEtran contains the IEEEeqnarray family of commands that can be used to
-% generate multiline equations as well as matrices, tables, etc., of high
-% quality.
-
-%\usepackage{eqparbox}
-% Also of notable interest is Scott Pakin's eqparbox package for creating
-% (automatically sized) equal width boxes - aka "natural width parboxes".
-% Available at:
-% http://www.ctan.org/tex-archive/macros/latex/contrib/eqparbox/
-
-% *** SUBFIGURE PACKAGES ***
-%\usepackage[tight,footnotesize]{subfigure}
-% subfigure.sty was written by Steven Douglas Cochran. This package makes it
-% easy to put subfigures in your figures. e.g., "Figure 1a and 1b". For IEEE
-% work, it is a good idea to load it with the tight package option to reduce
-% the amount of white space around the subfigures. subfigure.sty is already
-% installed on most LaTeX systems. The latest version and documentation can
-% be obtained at:
-% http://www.ctan.org/tex-archive/obsolete/macros/latex/contrib/subfigure/
-% subfigure.sty has been superceeded by subfig.sty.
-
-%\usepackage[caption=false]{caption}
-%\usepackage[font=footnotesize]{subfig}
-% subfig.sty, also written by Steven Douglas Cochran, is the modern
-% replacement for subfigure.sty. However, subfig.sty requires and
-% automatically loads Axel Sommerfeldt's caption.sty which will override
-% IEEEtran.cls handling of captions and this will result in nonIEEE style
-% figure/table captions. To prevent this problem, be sure and preload
-% caption.sty with its "caption=false" package option. This is will preserve
-% IEEEtran.cls handing of captions. Version 1.3 (2005/06/28) and later 
-% (recommended due to many improvements over 1.2) of subfig.sty supports
-% the caption=false option directly:
-%\usepackage[caption=false,font=footnotesize]{subfig}
-%
-% The latest version and documentation can be obtained at:
-% http://www.ctan.org/tex-archive/macros/latex/contrib/subfig/
-% The latest version and documentation of caption.sty can be obtained at:
-% http://www.ctan.org/tex-archive/macros/latex/contrib/caption/
-
-% *** FLOAT PACKAGES ***
-%
-%\usepackage{fixltx2e}
-% fixltx2e, the successor to the earlier fix2col.sty, was written by
-% Frank Mittelbach and David Carlisle. This package corrects a few problems
-% in the LaTeX2e kernel, the most notable of which is that in current
-% LaTeX2e releases, the ordering of single and double column floats is not
-% guaranteed to be preserved. Thus, an unpatched LaTeX2e can allow a
-% single column figure to be placed prior to an earlier double column
-% figure. The latest version and documentation can be found at:
-% http://www.ctan.org/tex-archive/macros/latex/base/
-
-%\usepackage{stfloats}
-% stfloats.sty was written by Sigitas Tolusis. This package gives LaTeX2e
-% the ability to do double column floats at the bottom of the page as well
-% as the top. (e.g., "\begin{figure*}[!b]" is not normally possible in
-% LaTeX2e). It also provides a command:
-%\fnbelowfloat
-% to enable the placement of footnotes below bottom floats (the standard
-% LaTeX2e kernel puts them above bottom floats). This is an invasive package
-% which rewrites many portions of the LaTeX2e float routines. It may not work
-% with other packages that modify the LaTeX2e float routines. The latest
-% version and documentation can be obtained at:
-% http://www.ctan.org/tex-archive/macros/latex/contrib/sttools/
-% Documentation is contained in the stfloats.sty comments as well as in the
-% presfull.pdf file. Do not use the stfloats baselinefloat ability as IEEE
-% does not allow \baselineskip to stretch. Authors submitting work to the
-% IEEE should note that IEEE rarely uses double column equations and
-% that authors should try to avoid such use. Do not be tempted to use the
-% cuted.sty or midfloat.sty packages (also by Sigitas Tolusis) as IEEE does
-% not format its papers in such ways.
-
-% *** PDF, URL AND HYPERLINK PACKAGES ***
-%
 \usepackage{url}
-% url.sty was written by Donald Arseneau. It provides better support for
-% handling and breaking URLs. url.sty is already installed on most LaTeX
-% systems. The latest version can be obtained at:
-% http://www.ctan.org/tex-archive/macros/latex/contrib/misc/
-% Read the url.sty source comments for usage information. Basically,
-% \url{my_url_here}.
 
@@
 
-% *** Do not adjust lengths that control margins, column widths, etc. ***
-% *** Do not use packages that alter fonts (such as pslatex).         ***
-% There should be no need to do such things with IEEEtran.cls V1.6 and later.
-% (Unless specifically asked to do so by the journal or conference you plan
-% to submit to, of course. )
+\usepackage{textpos}
+\begin{document}
 
-% correct bad hyphenation here
-\hyphenation{op-tical net-works semi-conduc-tor}
-\input{macros}
+\conferenceinfo{-}{-} 
+\copyrightyear{-} 
+\copyrightdata{[to be supplied]}
 
-\begin{document}
-%
-% paper title
-% can use linebreaks \\ within to get better formatting as desired
-\title{{Ace}: An Actively-Typed Language and Compilation Environment for High-Performance Computing}
+%\titlebanner{{\tt \textcolor{Red}{{\small Under Review -- distribute within CMU only.}}}}        % These are ignored unless
+%\preprintfooter{Distribute within CMU only.}   % 'preprint' option specified.
 
-% author names and affiliations
-% use a multiple column layout for up to two different
-% affiliations
-\author{\alignauthor
-Cyrus Omar and Jonathan Aldrich\\
-       \affaddr{Carnegie Mellon University, Pittsburgh, PA}\\
-       \email{\{comar,aldrich\}@cs.cmu.edu}
-}
+\newcommand{\Ace}{\textsf{Ace}}
 
-% conference papers do not typically use \thanks and this command
-% is locked out in conference mode. If really needed, such as for
-% the acknowledgment of grants, issue a \IEEEoverridecommandlockouts
-% after \documentclass
+\title{\Ace: Growing a Statically-Typed Language Inside a Python}
 
-% for over three affiliations, or if they all won't fit within the width
-% of the page, use this alternative format:
-% 
-%\author{\IEEEauthorblockN{Michael Shell\IEEEauthorrefmark{1},
-%Homer Simpson\IEEEauthorrefmark{2},
-%James Kirk\IEEEauthorrefmark{3}, 
-%Montgomery Scott\IEEEauthorrefmark{3} and
-%Eldon Tyrell\IEEEauthorrefmark{4}}
-%\IEEEauthorblockA{\IEEEauthorrefmark{1}School of Electrical and Computer Engineering\\
-%Georgia Institute of Technology,
-%Atlanta, Georgia 30332--0250\\ Email: see http://www.michaelshell.org/contact.html}
-%\IEEEauthorblockA{\IEEEauthorrefmark{2}Twentieth Century Fox, Springfield, USA\\
-%Email: homer@thesimpsons.com}
-%\IEEEauthorblockA{\IEEEauthorrefmark{3}Starfleet Academy, San Francisco, California 96678-2391\\
-%Telephone: (800) 555--1212, Fax: (888) 555--1212}
-%\IEEEauthorblockA{\IEEEauthorrefmark{4}Tyrell Inc., 123 Replicant Street, Los Angeles, California 90210--4321}}
+\authorinfo{~}{~}{~}
+%\authorinfo{Cyrus Omar\and Jonathan Aldrich}
+ %          {School of Computer Science\\
+  %          Carnegie Mellon University}
+   %        {\{comar, jonathan.aldrich\}@cs.cmu.edu}
 
-% use for special paper notices
-%\IEEEspecialpapernotice{(Invited Paper)}
-
-% make the title area
 \maketitle
-
 \begin{abstract}
@@
 We validate our overall design by considering a case study of a simulation framework enabling the modular specification and efficient execution of ensembles of neural simulations across a cluster of GPUs.
-
-%in the context of the applications and architectures that the language may need to target.
-%
-%Programming languages targeted toward end-users in high-performance computing must consider a number of issues in addition to performance.
-Typically, researchers have focused verifiability, portability and ease-of-use, and the right balance often depends on the particular uses that they are put toward.
-%
-%Ace %, evaluates it against several design and adoption criteria, and describes several use cases as well as a realistic case study involving parallelizing neurobiological circuit simulations on clusters of GPUs.
-%is an extensible statically-typed programming language embedded within the ubiquitous scripting language Python.
-Python serves as a {metalanguage}, enabling programmatic control over several aspects of language specification and compilation.
-%We demonstrate the power of this design for HPC by defining all of the primitives of the OpenCL kernel language as a library of user-defined types in Ace, then extending this core with several higher-level first-class abstractions.
-The target of translation, called the \emph{backend}, is user-defined as well, allowing us to implement the semantics of these primitives by direct translation to OpenCL, CUDA or C99.
-%
-%Although Ace is built around a static type discipline, it supports a novel form of generic programming based on type propagation that, together with a form of extensible local type inference, eliminates most of the syntactic overhead usually associated with statically-typed languages.
-A companion host API allows type information from \verb|numpy| arrays to directly propagate into Ace kernels, supporting seamless adoption of Ace into current scientific workflows.
-%
 \end{abstract}
 
-%\begin{IEEEkeywords}
-%extensible programming languages, heterogeneous computing, type systems
-%\end{IEEEkeywords}
-
-% For peer review papers, you can put extra information on the cover
-% page as needed:
-% \ifCLASSOPTIONpeerreview
-% \begin{center} \bfseries EDICS Category: 3-BBND \end{center}
-% \fi
-%
-% For peerreview papers, this IEEEtran command inserts a page break and
-% creates the second title. It will be ignored for other modes.
-%\IEEEpeerreviewmaketitle
-
+%\category{D.3.2}{Programming Languages}{Language Classifications}[Extensible Languages]
+%\category{D.3.4}{Programming Languages}{Processors}[Compilers]
+%\category{F.3.1}{Logics \& Meanings of Programs}{Specifying and Verifying and Reasoning about Programs}[Specification Techniques]
+%\keywords
+%type-level computation, typed compilation
 \section{Introduction}
-
 Computer-aided simulation and data analysis techniques have transformed science and engineering.
@@
 We hope that Ace will be developed further by the community to strengthen the foundations upon which new abstractions are implemented and deployed into the HPC professional end-user community.
-\newpage
-%\section{Availability}
-%Ace is available under the LGPL license and is developed openly and collaboratively using the popular Github platform at \url{https://github.com/cyrus-/ace}.
-Documentation, examples and other learning materials will be available at \url{http://acelang.org/}.
-{\color{red}(by the time of the conference)}
-%
-%\section{Acknowledgments}
-%CO was funded by the DOE Computational Science Graduate Fellowship under grant number DE-FG02-97ER25308. NF was funded by ...
-%
-% An example of a floating figure using the graphicx package.
-% Note that \label must occur AFTER (or within) \caption.
-% For figures, \caption should occur after the \includegraphics.
-% Note that IEEEtran v1.7 and later has special internal code that
-% is designed to preserve the operation of \label within \caption
-% even when the captionsoff option is in effect. However, because
-% of issues like this, it may be the safest practice to put all your
-% \label just after \caption rather than within \caption{}.
-%
-% Reminder: the "draftcls" or "draftclsnofoot", not "draft", class
-% option should be used if it is desired that the figures are to be
-% displayed while in draft mode.
-%
-%\begin{figure}[!t]
-%\centering
-%\includegraphics[width=2.5in]{myfigure}
-% where an .eps filename suffix will be assumed under latex, 
-% and a .pdf suffix will be assumed for pdflatex; or what has been declared
-% via \DeclareGraphicsExtensions.
-%\caption{Simulation Results}
-%\label{fig_sim}
-%\end{figure}
 
-% Note that IEEE typically puts floats only at the top, even when this
-% results in a large percentage of a column being occupied by floats.
+%\acks
+%The author is grateful to Jonathan Aldrich, Robert Bocchino and anonymous referees for their useful suggestions.
+This work was funded by the DOE Computational Science Graduate Fellowship under grant number DE-FG02-97ER25308.
+%Acknowledgments, if needed.
 
-% An example of a double column floating figure using two subfigures.
-% (The subfig.sty package must be loaded for this to work.)
-% The subfigure \label commands are set within each subfloat command, the
-% \label for the overall figure must come after \caption.
-% \hfil must be used as a separator to get equal spacing.
-% The subfigure.sty package works much the same way, except \subfigure is
-% used instead of \subfloat.
-%
-%\begin{figure*}[!t]
-%\centerline{\subfloat[Case I]\includegraphics[width=2.5in]{subfigcase1}%
-%\label{fig_first_case}}
-%\hfil
-%\subfloat[Case II]{\includegraphics[width=2.5in]{subfigcase2}%
-%\label{fig_second_case}}}
-%\caption{Simulation results}
-%\label{fig_sim}
-%\end{figure*}
-%
-% Note that often IEEE papers with subfigures do not employ subfigure
-% captions (using the optional argument to \subfloat), but instead will
-% reference/describe all of them (a), (b), etc., within the main caption.
+% We recommend abbrvnat bibliography style.
 
-% An example of a floating table. Note that, for IEEE style tables, the 
-% \caption command should come BEFORE the table. Table text will default to
-% \footnotesize as IEEE normally uses this smaller font for tables.
-% The \label must come after \caption as always.
-%
-%\begin{table}[!t]
-%% increase table row spacing, adjust to taste
-%\renewcommand{\arraystretch}{1.3}
-% if using array.sty, it might be a good idea to tweak the value of
-% \extrarowheight as needed to properly center the text within the cells
-%\caption{An Example of a Table}
-%\label{table_example}
-%\centering
-%% Some packages, such as MDW tools, offer better commands for making tables
-%% than the plain LaTeX2e tabular which is used here.
-%\begin{tabular}{|c||c|}
-%\hline
-%One & Two\\
-%\hline
-%Three & Four\\
-%\hline
-%\end{tabular}
-%\end{table}
+\bibliographystyle{abbrvnat}
 
-% Note that IEEE does not put floats in the very first column - or typically
-% anywhere on the first page for that matter. Also, in-text middle ("here")
-% positioning is not used. Most IEEE journals/conferences use top floats
-% exclusively. Note that, LaTeX2e, unlike IEEE journals/conferences, places
-% footnotes above bottom floats. This can be corrected via the \fnbelowfloat
-% command of the stfloats package.
+% The bibliography should be embedded for final submission.
 
-% conference papers do not normally have an appendix
+\bibliography{../research}
+%\softraggedright
+%P. Q. Smith, and X. Y. Jones. ...reference text...
 
-% use section* for acknowledgement
-% trigger a \newpage just before the given reference
-% number - used to balance the columns on the last page
-% adjust value as needed - may need to be readjusted if
-% the document is modified later
-%\IEEEtriggeratref{8}
-% The "triggered" command can be changed if desired:
-%\IEEEtriggercmd{\enlargethispage{-5in}}
-
-% references section
-
-% can use a bibliography generated by BibTeX as a .bbl file
-% BibTeX documentation can be easily obtained at:
-% http://www.ctan.org/tex-archive/biblio/bibtex/contrib/doc/
-% The IEEEtran BibTeX style support page is at:
-% http://www.michaelshell.org/tex/ieeetran/bibtex/
-%\bibliographystyle{IEEEtran}
-% argument is your BibTeX string definitions and bibliography database(s)
-%\bibliography{IEEEabrv,../bib/paper}
-%
-% <OR> manually copy in the resultant .bbl file
-% set second argument of \begin to the number of references
-% (used to reserve space for the reference number labels box)
-\bibliographystyle{IEEEtran}
-\bibliography{../research}
-%
-%\begin{thebibliography}{1}
-%
-%\bibitem{IEEEhowto:kopka}
-%H.~Kopka and P.~W. Daly, \emph{A Guide to \LaTeX}, 3rd~ed.\hskip 1em plus
-%  0.5em minus 0.4em\relax Harlow, England: Addison-Wesley, 1999.
-%
-%\end{thebibliography}
-
-% that's all folks
 \end{document}

### 2013-11-07 ace-pldi14/listing10.py

@@ -1,0 +1,5 @@
+sim = Simulation(None, n_timesteps=10000)
+neurons = ReducedLIF(sim, count=N, tau=20.0)
+e_synapse = ExponentialSynapse(neurons, 'ge',
+    tau=5.0, reversal=60.0)
+probe = StateVariableProbeCopyback(e_synapse)

### 2013-11-07 ace-pldi14/listing3.py

@@ -1,0 +1,6 @@
+from ace.OpenCL import OpenCL, get_global_id
+
+@OpenCL.fn
+def map(input, output, f):
+    gid = get_global_id(0)
+    output[gid] = f(input[gid])

### 2013-11-07 ace-pldi14/listing4.py

@@ -1,0 +1,10 @@
+from listing3 import map
+from ace.OpenCL import gptr, double, int
+
+@OpenCL.fn
+def add5(x):
+    return x + 5
+
+D = gptr(double); I = gptr(int); A = add5.ace_type
+map_add5_dbl = map.compile(D, D, A)
+map_add5_int = map.compile(I, I, A)

### 2013-11-07 ace-pldi14/listing9.py

@@ -1,0 +1,25 @@
+import ace
+
+class PtrType(ace.Type):
+  def __init__(self, target_type, addr_space):
+    self.target_type = target_type
+    self.addr_space = addr_space
+        
+  def resolve_Subscript(self, context, node):
+    slice_type = context.resolve(node.slice)
+    if isinstance(slice_type, IntegerType):
+      return self.target_type
+    else:
+      raise TypeError('<error message>', node)
+       
+  def translate_Subscript(self, context, node):
+    value = context.translate(node.value)
+    slice = context.translate(node.slice)
+    return ace.copy_node(node, 
+      value=value, slice=slice,    
+      code=value.code + '[' + slice.code + ']')      
+  
+  # ...
+  
+def gptr(target_type):
+    return PtrType(target_type, "__global")

### 2013-11-07 ace-pldi14/ace-pldi14.tex
Began revamping PLDI14 abstract

@@
 \begin{abstract}
-Researchers developing languages and abstractions for high-performance computing must consider a number of design criteria, including performance, verifiability, portability and ease-of-use.
+Evidence suggests that programmers are unwilling to adopt new languages to gain access to language-integrated abstractions, even when they agree that these abstractions might be valuable if all else were held equal.
+This suggests that language designers should give greater consideration to two related design criteria: \emph{compatibility} with existing languages, tools and ecosystems, and \emph{extensibility}, so that adopting a new language-integrated abstraction does not require adopting a new language.
+
+In this paper, we introduce \Ace, a language compatible with the tools and infrastructure developed for the world's most widely-used dynamically-typed language, Python. 
+%Researchers developing languages and abstractions for high-performance computing must consider a number of design criteria, including performance, verifiability, portability and ease-of-use.
 Despite the deficiencies of legacy tools and the availability of seemingly superior options, end-users have been reluctant to adopt new language-based abstractions.

### 2013-11-07 ace-pldi14/ace-pldi14.tex
Finished abstract of PLDI14

@@
 
+\usepackage{todonotes}
+
 \usepackage{placeins}
@@
 \begin{abstract}
-Evidence suggests that programmers are unwilling to adopt new languages to gain access to language-integrated abstractions, even when they agree that these abstractions might be valuable if all else were held equal.
-This suggests that language designers should give greater consideration to two related design criteria: \emph{compatibility} with existing languages, tools and ecosystems, and \emph{extensibility}, so that adopting a new language-integrated abstraction does not require adopting a new language.
+Evidence suggests that programmers are reluctant to adopt new programming languages to gain access to new abstractions, even when they agree that these abstractions could be valuable for their work.
+This suggests a need for languages designed to satisfy two related design criteria: \emph{compatibility} with existing languages, tools and ecosystems; and \emph{extensibility}, so that adopting new abstractions does not require adopting a new language.
 
-In this paper, we introduce \Ace, a language compatible with the tools and infrastructure developed for the world's most widely-used dynamically-typed language, Python. 
+In this paper, we introduce \Ace, a language compatible with the tools and infrastructure developed for Python, among the most  widely-adopted dynamically-typed language today.
+While Python and other similar languages are considered satisfactory for small-scale scripting tasks, \Ace~is designed for situations where correctness and performance are particularly important, offering a  static type system and more granular control over compilation.
+Unlike most statically-typed languages, however, the \Ace~type system can be extended from within.
+The extension mechanism, called \emph{active typechecking and translation}, guarantees that extensions cannot interfere  with one another by restricting the scope of extension logic, which must be associated with a type, around expressions of that type.
+We show that this mechanism remains powerful enough to express the static and dynamic semantics of several common functional, object-oriented and parallel programming primitives, as well as varied domain-specific abstractions, all from within libraries.
+Ace can be used from the shell or interactively from within Python\todo{take this sentence out?}.
 %Researchers developing languages and abstractions for high-performance computing must consider a number of design criteria, including performance, verifiability, portability and ease-of-use.

### 2013-11-08 ace-pldi14/ace-pldi14.tex
Tweaked PLDI14 abstract and sections

@@
 In this paper, we introduce \Ace, a language compatible with the tools and infrastructure developed for Python, among the most  widely-adopted dynamically-typed language today.
-While Python and other similar languages are considered satisfactory for small-scale scripting tasks, \Ace~is designed for situations where correctness and performance are particularly important, offering a  static type system and more granular control over compilation.
-Unlike most statically-typed languages, however, the \Ace~type system can be extended from within.
-The extension mechanism, called \emph{active typechecking and translation}, guarantees that extensions cannot interfere  with one another by restricting the scope of extension logic, which must be associated with a type, around expressions of that type.
-We show that this mechanism remains powerful enough to express the static and dynamic semantics of several common functional, object-oriented and parallel programming primitives, as well as varied domain-specific abstractions, all from within libraries.
-Ace can be used from the shell or interactively from within Python\todo{take this sentence out?}.
+While Python and other similar languages have proven satisfactory for small-scale scripting tasks, \Ace~is designed for situations where correctness and performance are particularly important, offering static typechecking and more granular control over compilation.
+Unlike most statically-typed languages, \Ace's type system can be extended from within.
+The extension mechanism guarantees that extensions cannot interfere with one another by organizing extension logic around types.
+We show that such \emph{active types} can express the static and dynamic semantics of several primitive functional, object-oriented and parallel programming abstractions, as well as domain-specific abstractions, all from within libraries.
+%Ace can be used from the shell or interactively from within Python\todo{take this sentence out?}.
 %Researchers developing languages and abstractions for high-performance computing must consider a number of design criteria, including performance, verifiability, portability and ease-of-use.
@@
 Note that this example would also work correctly if the assignments to \verb|y| were replaced with \verb|return| statements (in other words, the return value of a function is treated as an assignable for the purpose of type inference).
-\subsection{Annotation and Extension Inference}
-In addition to type annotations, OpenCL normally asks for additional annotations in a number of other situations.
-Users can annotate functions that meet certain requirements  with the \verb|__kernel| attribute, indicating that they are callable from the host.
-The \verb|OpenCL| backend can check these and add this annotation automatically.
-Several types (notably, \verb|double|) and specialized primitive functions require that an OpenCL extension be enabled via a \verb|#pragma|.
-The OpenCL backend automatically detects many of these cases as well, adding the appropriate \verb|#pragma| declaration.
-An example of this can be seen in Listing \ref{mapout}, where the use of the \verb|double| type triggers the insertion of the \verb|cl_khr_f64| extension.
 
@@
 \subsubsection{Parallel Programming}
+\paragraph{OpenCL}
+\subparagraph{Annotation and Extension Inference}
+In addition to type annotations, OpenCL normally asks for additional annotations in a number of other situations.
+Users can annotate functions that meet certain requirements  with the \verb|__kernel| attribute, indicating that they are callable from the host.
+The \verb|OpenCL| backend can check these and add this annotation automatically.
+Several types (notably, \verb|double|) and specialized primitive functions require that an OpenCL extension be enabled via a \verb|#pragma|.
+The OpenCL backend automatically detects many of these cases as well, adding the appropriate \verb|#pragma| declaration.
+An example of this can be seen in Listing \ref{mapout}, where the use of the \verb|double| type triggers the insertion of the \verb|cl_khr_f64| extension.
+
 \paragraph{Partitioned Global Address Spaces}
@@
 
-\section{Case Study}\label{casestudy}
-An important criteria that practitioners use to evaluate a language or abstraction is whether significant case studies have been conducted with it, sometimes called \textbf{social proof} \cite{basili2008understanding}.
-In this section, we briefly (due to space constraints) show an application of the mode of use discussed above to support a modular, high-performance scientific simulation library.
-This library was used to simulate  thousands of parallel realizations of a stochastic spiking neurobiological circuit.
-The realizations were distributed over a cluster of GPUs using an existing Python library supporting remote procedure calls.
-This relied critically on Ace's representation of types to determine the memory usage of each realization automatically, and thus distribute realizations and allocate memory on each node appropriately.
-It also used metaprogramming features of Ace to generate kernels and specialize them to each node the simulation was running on.
-
 \subsection{Metaprogramming in Ace}
@@
 This transformation as well as some others are distributed in \verb|ace.astx| for convenience.
-
-\subsection{Background}
-A neural circuit can be modeled as a network of coupled differential equations, where each node corresponds to a single neuron or a small population.
-Each node has temporal dynamics given by one or more ordinary differential equations.
-Single simulations can contain from hundreds to tens of millions of neurons each, depending on the specific problem being studied.
-In some cases, such as when studying the effects of noise on network dynamics or to sweep a parameter space, hundreds or thousands of realizations must be generated.
-In these cases, care must be taken to only probe the simulation for relevant data and process portions of it as the simulation progresses, because the amount of data generated is often too large to store in its entirety.
-
-The problem we discuss here involved studying a problem that required running up to 1,000 realizations of a network of between 4,000 and 10,000 stochastic neurons each.
-An initial solution to this problem used the Brian framework, written in Python, to conduct these simulations on a CPU.
-Brian was selected because it allowed the structure of the  simulation to be specified in a modular and straightforward manner.
-This solution required between 60 and 70 minutes to conduct the simulations and up to 8 hours to analyze the data each time a parameter of the simulation was modified.
-
-Unsatisfied with the performance of this approach, the author developed an accelerated variant of the simulation using C++ and CUDA.
-Although this produced significant speedups, reducing the time for a simulation by a factor of 40 and the runtime of the slowest analyses by a factor of 200, the overall workflow was also significantly disrupted.
-In order to support the many variants of models, parameter sets, and probing protocols, C preprocessor flags were necessary to selectively include or exclude code snippets.
-This quickly led to an incomprehensible and difficult to maintain file structure.
-Moreover, much of the subsequent data analysis and visualization was conducted using Python, so marshalling the relevant data between processes also became an issue.
-Doing so during a simulation was not attempted.
-
-\subsection{The {\sf cl.egans} Simulation Library}
-
-\begin{codelisting}
-\lstinputlisting{listing10.py}
-\caption{\texttt{[listing10.py]} An example of a nested simulation tree, showing that specifying a simulation is both simple and modular.
-The first argument to the constructor specifies each node's parent.}
-\label{spec}
-\end{codelisting}
-
-\begin{codelisting}
-\lstinputlisting{listing11.py}
-\caption{\texttt{[listing11.py]} An example of a hook that inserts code and also inserts new, nested hooks for downstream simulation nodes  below that.}
-\label{impl}
-\end{codelisting}
-
-In order to eliminate these issues while retaining the performance profile of the GPU-accelerated code, the project was ported to Ace.
-Rather than using preprocessor directives to control the code contained in the final GPU kernels used to execute the simulation and data analyses, the group was able to develop a more modular  library called {\sf cl.egans}\footnote{...after {\it c.
-elegans}, a model organism in neuroscience} based on the language's compile-time code generation mechanism and Python and OpenCL bindings.
-
-{\sf cl.egans} leverages Python's object-oriented features to enable modular, hierarchical simulation specifications.
-For example, Figure \ref{spec} shows an example where a neuron model (\verb|ReducedLIF|) is added to the root of the simulation, a synapse model (\verb|ExponentialSynapse|) is then added to it, and its conductance is probed by adding a probe model as a child of the synapse model.
-Interleaved analyses are specified as nodes as well (not shown).
-
-Implementations of these classes do not evaluate the simulation logic directly, but rather contain methods that generate Ace source code for insertion at various points, called {\it hooks}, in the final simulation kernel.
-The hook that code is inserted into is determined by the method name, and code can be inserted into any hook defined anywhere upstream in the simulation tree.
-New hooks can also be defined in these methods and these become available for use by child nodes.
-Figure \ref{impl} shows an example of a class that inserts code in the \verb|model_code| hook and defines several new hooks.
-This protocol is closely related to the notion of {\it frame-oriented programming}.
-Although highly modular, this strategy avoids the performance penalties associated with standard object-oriented methodologies via code generation.
-
-Compared to a similar protocol targeting OpenCL directly, the required code generation logic is significantly simpler because it enables classes like \verb|StateVariable| to be written generically for all types of state variables, without carrying extra parameters and {\it ad hoc} logic to extract and compute the result types of generated expressions.
-Moreover, because types are first-class objects in the metalanguage, they can be examined during the memory allocation step to enable features like fully-automatic parallelization of multiple realizations across one or more devices, a major feature of {\sf cl.egans} that competing frameworks cannot easily offer.
-Moreover, memory allocation is significantly simplified by the availability of typing information at compile-time.
-The amount of memory needed by a simulation can be calculated statically and thus distributing the simulation over a cluster simply requires determining the available memory on each device in the cluster.
-
-Once the kernel has been generated and memory has been allocated, the simulation can be executed directly  using the bindings above.
-The results of this simulation are available to the Python code during and following the simulation and can be visualized and further analyzed using standard tools.
-Once the computations is complete, the  garbage collector is able to handle deallocation of GPU memory automatically (a feature of the underlying \verb|pyopencl| library \cite{klockner2011pycuda}.)
-
-Using this Ace-based framework, the benefits of the Brian-based workflow were recovered without the  corresponding decrease in performance relative to the previous CUDA-based solution, leading ultimately to a satisfying solution.
 

### 2013-11-13 ace-pldi14/ace-pldi14.tex
Reworking of Ace paper for PLDI14

@@
 \usepackage{listings}
+
+    \makeatletter
+%
+% \btIfInRange{number}{range list}{TRUE}{FALSE}
+%
+% Test if int number <number> is element of a (comma separated) list of ranges
+% (such as: {1,3-5,7,10-12,14}) and processes <TRUE> or <FALSE> respectively
+%
+        \newcount\bt@rangea
+        \newcount\bt@rangeb
+
+        \newcommand\btIfInRange[2]{%
+            \global\let\bt@inrange\@secondoftwo%
+            \edef\bt@rangelist{#2}%
+            \foreach \range in \bt@rangelist {%
+                \afterassignment\bt@getrangeb%
+                \bt@rangea=0\range\relax%
+                \pgfmathtruncatemacro\result{ ( #1 >= \bt@rangea) && (#1 <= \bt@rangeb) }%
+                \ifnum\result=1\relax%
+                    \breakforeach%
+                    \global\let\bt@inrange\@firstoftwo%
+                \fi%
+            }%
+            \bt@inrange%
+        }
+
+        \newcommand\bt@getrangeb{%
+            \@ifnextchar\relax%
+            {\bt@rangeb=\bt@rangea}%
+            {\@getrangeb}%
+        }
+
+        \def\@getrangeb-#1\relax{%
+            \ifx\relax#1\relax%
+                \bt@rangeb=100000%   \maxdimen is too large for pgfmath
+            \else%
+                \bt@rangeb=#1\relax%
+            \fi%
+        }
+
+%
+% \btLstHL{range list}
+%
+        \newcommand{\btLstHL}[1]{%
+            \btIfInRange{\value{lstnumber}}{#1}%
+            {\color{black!10}}%
+            {\def\lst@linebgrd}%
+        }%
+
+%
+% \btInputEmph[listing options]{range list}{file name}
+%
+        \newcommand{\btLstInputEmph}[3][\empty]{%
+            \lstset{%
+                linebackgroundcolor=\btLstHL{#2}%
+                \lstinputlisting{#3}%
+            }% \only
+        }
+
+% Patch line number key to call line background macro
+        \lst@Key{numbers}{none}{%
+            \def\lst@PlaceNumber{\lst@linebgrd}%
+            \lstKV@SwitchCases{#1}{%
+                none&\\%
+                left&\def\lst@PlaceNumber{\llap{\normalfont
+                \lst@numberstyle{\thelstnumber}\kern\lst@numbersep}\lst@linebgrd}\\%
+                right&\def\lst@PlaceNumber{\rlap{\normalfont
+                \kern\linewidth \kern\lst@numbersep
+                \lst@numberstyle{\thelstnumber}}\lst@linebgrd}%
+            }{%
+                \PackageError{Listings}{Numbers #1 unknown}\@ehc%
+            }%
+        }
+
+% New keys
+        \lst@Key{linebackgroundcolor}{}{%
+            \def\lst@linebgrdcolor{#1}%
+        }
+        \lst@Key{linebackgroundsep}{0pt}{%
+            \def\lst@linebgrdsep{#1}%
+        }
+        \lst@Key{linebackgroundwidth}{\linewidth}{%
+            \def\lst@linebgrdwidth{#1}%
+        }
+        \lst@Key{linebackgroundheight}{\ht\strutbox}{%
+            \def\lst@linebgrdheight{#1}%
+        }
+        \lst@Key{linebackgrounddepth}{\dp\strutbox}{%
+            \def\lst@linebgrddepth{#1}%
+        }
+        \lst@Key{linebackgroundcmd}{\color@block}{%
+            \def\lst@linebgrdcmd{#1}%
+        }
+
+% Line Background macro
+        \newcommand{\lst@linebgrd}{%
+            \ifx\lst@linebgrdcolor\empty\else
+                \rlap{%
+                    \lst@basicstyle
+                    \color{-.}% By default use the opposite (`-`) of the current color (`.`) as background
+                    \lst@linebgrdcolor{%
+                        \kern-\dimexpr\lst@linebgrdsep\relax%
+                        \lst@linebgrdcmd{\lst@linebgrdwidth}{\lst@linebgrdheight}{\lst@linebgrddepth}%
+                    }%
+                }%
+            \fi
+        }
+
+    \makeatother
+
 \lstset{
@@
     language={}, 
+    numbers=left,
+    numberstyle=\scriptsize\color{light-gray}\textsf,
     moredelim=**[is][\color{blue}\bf\ttfamily]{`}{`},
@@
 	language=C++,
-	morekeywords={kernel, __kernel, global, __global, size_t, get_global_id, sin}
+	morekeywords={kernel, __kernel, global, __global, size_t, get_global_id, sin, printf, int2}
 }
@@
 \floatname{codelisting}{Listing}
+\setlength{\floatsep}{10pt}
+\setlength{\textfloatsep}{10pt}
 
@@
 \usepackage{textpos}
+
+\renewcommand\topfraction{0.85}
+\renewcommand\bottomfraction{0.85}
+\renewcommand\textfraction{0.1}
+\renewcommand\floatpagefraction{0.85}
+
 \begin{document}
@@
 \begin{abstract}
-Evidence suggests that programmers are reluctant to adopt new programming languages to gain access to new abstractions, even when they agree that these abstractions could be valuable for their work.
-This suggests a need for languages designed to satisfy two related design criteria: \emph{compatibility} with existing languages, tools and ecosystems; and \emph{extensibility}, so that adopting new abstractions does not require adopting a new language.
+%Evidence suggests that programmers are reluctant to adopt new languages to
+%gain access to new abstractions, even when they agree that these abstractions
+%could be valuable. This suggests a need for languages that satisfy two key
+%growth criteria: compatibility with existing programming ecosystems; and
+%internal extensibility, so that new abstractions do not require new
+%languages.
+%
+%We introduce Ace, a language compatible with the popular Python ecosystem.
+%While Python, like other similar languages, is satisfactory for simple
+%scripting, Ace is designed for more complex situations where static
+%typechecking is beneficial. Unlike most statically-typed languages, Ace's type
+%system can be extended from within, by associating compile-time functions with
+%type definitions. The compiler selectively invokes these according to a
+%type-directed protocol, avoiding interference problems. Despite these design
+%constraints, active types in Ace can express the static and dynamic semantics
+%of a range of functional, object-oriented, parallel and domain-specific
+%abstractions, all within libraries.
 
-In this paper, we introduce \Ace, a language compatible with the tools and infrastructure developed for Python, among the most  widely-adopted dynamically-typed language today.
-While Python and other similar languages have proven satisfactory for small-scale scripting tasks, \Ace~is designed for situations where correctness and performance are particularly important, offering static typechecking and more granular control over compilation.
-Unlike most statically-typed languages, \Ace's type system can be extended from within.
-The extension mechanism guarantees that extensions cannot interfere with one another by organizing extension logic around types.
-We show that such \emph{active types} can express the static and dynamic semantics of several primitive functional, object-oriented and parallel programming abstractions, as well as domain-specific abstractions, all from within libraries.
+Evidence suggests that programmers are reluctant to adopt new programming languages to gain access to new abstractions, even when they agree that these abstractions would be valuable to them.
+This suggests a need for languages that are  \emph{compatible} with existing languages, tools and infrastructure and \emph{internally extensible}, so that adopting a new  primitive abstraction requires only importing a library in the usual way.
+
+In this paper, we introduce Ace, a language compatible with tools and infrastructure developed for Python, one of the most  widely-adopted dynamically-typed languages today.
+While Python, like other similar languages, was designed for simple scripting tasks, Ace is designed for more complex situations where static typechecking and programmatic control over compilation may be beneficial.
+Unlike most statically-typed languages, however, Ace's type system and semantics can be extended from within by novel mechanisms that avoid key interference issues faced by previous mechanisms.
+We show that these can be used to implement a range of statically-typed functional, object-oriented, parallel, low-level and domain-specific abstractions, as well as safe interoperability layers with existing languages, as orthogonal libraries.
 %Ace can be used from the shell or interactively from within Python\todo{take this sentence out?}.
@@
 \section{Introduction}
-Computer-aided simulation and data analysis techniques have transformed science and engineering.
-Surveys show that scientists and engineers now spend up to 40\% of their time writing software \cite{howison2011scientific, hannay2009scientists}.
-Most of this software targets desktop hardware, while about 20\% of scientists also target either local clusters or super\-computers for more numerically-intensive computations \cite{hannay2009scientists}.
-To fully harness the power of these platforms, however, these so-called {\em professional end-user developers} \cite{segal2007some} must increasingly write parallel programs.
+Asking programmers to import a new library is far more practical than asking them to  adopt a new programming language.
+Indeed, recent empirical studies underscore  the difficulties of driving new languages into adoption, finding that extrinsic factors like compatibility with existing codebases and libraries, team familiarity and tool support are at least as important as intrinsic factors \cite{Meyerovich:2013:EAP:2509136.2509515,chen05,nguyen2010survey}.
+As a result, many developers cannot use  abstractions they might prefer because these abstractions are only available in languages they cannot adopt \cite{Meyerovich:2012:SDR:2414721.2414724,Meyerovich:2013:EAP:2509136.2509515}.
+This issue was perhaps most succinctly expressed by a participant in a recent study by Basili et al.
+\cite{basili2008understanding} who stated ``I hate MPI, I hate C++.
+[But] if I had to choose again, I would probably choose the same.''
 
-Professional end-users today generally use dynamically-typed high-level languages like MATLAB, Python, R or Perl for tasks that are not performance-sensitive, such as small-scale data analysis and plotting \cite{nguyen2010survey}.
-For portions of their analyses where the performance overhead of dynamic type checking and automatic memory management is too high, they will typically call into code written in a statically-typed, low-level language, most commonly C or Fortran, that uses low-level parallel abstractions like pthreads and MPI \cite{4222616,basili2008understanding}.
-Unfortunately, these low-level languages and abstractions are notoriously difficult to use and automatic verification is intractable in general.
+Unfortunately, researchers and domain experts who design and develop potentially useful new abstractions can find it difficult to implement them in terms of the general-purpose abstraction mechanisms available in mainstream languages.
+This is particularly salient for abstractions that require support from the typechecker or compiler, such as those focused on correctness and performance, as well as those that  introduce more concise or natural notations for existing abstractions.
+For example, a recent controlled study comparing a new language, Habanero-Java (HJ), with a comparable library, \verb|java.util.concurrent|, found that the language-based approach was more concise, correct and easy-to-use, but concluded that the library-based approach was more practical outside the classroom because HJ introduced new constructs and keywords into Java, requiring the adoption of a new toolchain, which could lead to compatibility issues with plain Java code \cite{cave2010comparing}.
 
-Researchers and domain experts often respond to these challenges by proposing novel language features that aim to strike an intermediate balance between \textbf{performance}, \textbf{verifiability\-}, \textbf{portability} and \textbf{ease-of-use}.
-Unfortunately, professional end-users rarely adopt new languages.
-Indeed, many end-users have become skeptical that novel approaches that originate in the research community can be  practical.
-This viewpoint was perhaps most succinctly expressed by a participant in a recent study by Basili et al.
-\cite{basili2008understanding} who stated ``I hate MPI, I hate C++.
-[But] if I had to choose again, I would probably choose the same.'' Although it may seem paradoxical, the ubiquity of this sentiment demands direct examination by researchers proposing novel abstractions and languages for eventual use by professional end-users in HPC.
+Internally-extensible languages promise to reduce the need for new standalone languages by giving abstraction providers more direct control over a base language's syntax and semantics from within libraries.
+By using such a language, programmers  gain the ability to granularly import the primitive abstractions best suited to each part of their application or library.
+The research community thus gains the ability to more easily develop, deploy and evaluate new abstractions in the context of existing codebases, narrowing one of the gaps between research and practice \cite{basili2008understanding}.
 
-We suggest three mutually-related {design criteria} that, unlike those in bold above, many languages and language-integrated abstractions have failed to adequately consider: \textbf{continuity}, \textbf{extensibility} and \textbf{interoperability}.
-These criteria encompass the intuitions that new abstractions will not be adopted in a vacuum, that programming systems must  support change, and that interacting components of an application or workflow should be able to make use of different abstractions naturally and without the possibility of conflict arising at their interface boundaries.
-%that this is far from being a superfluous issue. These criteria can influence the core design of a language or abstraction.
+%Indeed, abstractions could compete on their individual merit, rather than  by the serious adoption barriers mentioned above.
 
-In this paper, we introduce Ace, a programming language targeting professional end-users as well as researchers across high-performance computing and related domains.
-Ace has followed a principled design methodology guided by this more comprehensive set of design criteria in order to avoid many of the issues that have hindered previous language-based approaches in high-performance computing.
-These criteria, this design methodology, and the novel mechanisms and designs developed to satisfy these criteria constitute the generalizable contributions of this paper.
-We also hope that Ace itself will be useful to the HPC community\footnote{\small Ace is openly available at http://acelang.org/.}, and we describe several use cases and an initial case study in order to preliminarily validate it's utility.
+Unfortunately, internally-extensible languages available today have several problems.
+First, an extension mechanism itself may require modifying a base language with constructs for defining, importing and using extensions.
+The extension mechanism is not itself a library, so it faces many of the same extrinsic issues as other new languages like HJ, leading to a ``bootstrapping'' problem.
+The extension mechanisms available today also have several intrinsic problems related to safety and expressiveness that require technical solutions before it would be appropriate to widely rely on them.
+We evaluate related work in Section \ref{related}.
 
- We begin in Section \ref{usage} with simple examples that show how Ace can be used for low-level GPU programming and introduce the fundamental decisions made in its design.
-We discuss the motivations, based on the criteria above, behind key design decisions, including the use of static typing, a fixed syntax based on Python, a novel type propagation and inference scheme, and an explicit phase separation between compile-time and run-time logic.
-Each of these can be seen in Listings 1 and 2, which we will discuss further below.
+This paper describes the design and implementation of Ace, an internally-extensible language designed considering both extrinsic and intrinsic criteria.
+To solve the bootstrapping problem, Ace is implemented entirely as a library within the popular Python programming language.
+Ace and Python share a common syntax and package system, allowing Ace to leverage its well-established tools and infrastructure directly.
+Python serves as the compile-time metalanguage for Ace, but Ace functions themselves do not operate according to Python's fixed dynamically-typed semantics  (cf.
+\cite{Politz:2013:PFM:2509136.2509536,python}).
+Instead, Ace has a statically-typed semantics that can be extended by users from within libraries.
 
-The examples in Section \ref{usage} are based on an internalization of the OpenCL kernel programming language \cite{opencl11} as a library.
-This library is built atop a novel extensibility mechanism that we call \emph{active typechecking and translation} (AT\&T) and detail in Section \ref{att}.
-This mechanism relies on the core idea of representing types as first-class objects at compile-time.
-We refer to these objects as \emph{active types} by analogy with \emph{active libraries} \cite{activelibraries} (see Section \ref{related}).
-Unlike prior approaches to extensibility where users globally modify the grammar or semantics of a language, and thus introduce conflicts between extensions, AT\&T guarantees that extensions are composable by construction by pairing new rules with new types and limiting their scope to expressions of that type.
+More specifically, each Ace function can be annotated with a base semantics that determines the meaning of simple expressions like literals and certain statements.
+The semantics of the remaining expressions and statements are governed by logic associated with the type of a designated subexpression.
+We call the user-defined base semantics \emph{active bases} and the types in Ace \emph{active types}, borrowing terminology from \emph{active libraries} (\cite{activelibraries}, see Sec.
+\ref{related}).
+Both are objects that can be defined and manipulated at compile-time using Python.
+An important consequence of this mechanism is that it permits \emph{compositional} reasoning -- active bases and active types govern only specific non-overlapping portions of a program.
+As a result, clients are able to import any combination of extensions with the confidence that link-time ambiguities cannot occur (unlike many previous approaches, as we discuss in Sec.
+\ref{related}).
 
-%Despite its simplicity and safety, this mechanism is flexible enough to define the entirety of the OpenCL kernel language (which is derived directly from C99) and, combined with Ace's type propagation scheme, substantially enhance it's ease-of-use.
-To demonstrate the flexibility of this mechanism beyond OpenCL, we continue in Section \ref{usecases} by outlining examples of other, higher-level parallel abstractions that can be cleanly implemented using AT\&T, including global address spaces, message passing and functional data parallelism.
-First-class support for each of these has required a new language in the past.
-AT\&T allows compile-time logic to be safely included within libraries, and thus these first-class abstractions can be coexist naturally within a single program or workflow.
+The \emph{target} of compilation is also user-defined.
+We will show examples of Ace targeting Python as well as OpenCL and CUDA, lower-level languages often used to program graphics hardware.
+An active base or type can support multiple target languages.
+Ace functions targeting a language with Python bindings can be called directly from Python scripts, with compilation occurring implicitly.
+In some cases, types can propagate from Python into Ace.
+We show how this can be used to streamline the kinds of interactive workflows that Python is often used for.
+Ace can also be used non-interactively from the shell, producing source files that can be further compiled and executed by external tools.
 
-Ace can be used as a standalone language via the \verb|acec| compiler and also as an interactive compilation environment.
-This mode of use, described in Section \ref{compenv}, uses just-in-time specialization to  integrate the widely-adopted \verb|numpy| library (including its internal type system) and OpenCL's host API (by a mechanism also available to other host APIs, such as CUDA's) with the Ace compiler itself to enable the direct invocation of Ace functions from within Python scripts with minimal overhead.
-Achieving this deep level of integration makes use of Ace's novel type representation as well.
-
-To demonstrate the utility of Ace for scientific workloads, we describe in Section \ref{casestudy} a case study where Ace was used in this mode to develop a scientific simulation framework.
-This framework has been used to  specify and efficiently execute thousands of realizations of a large stochastic neural circuit model on clusters of GPUs, achieving the same performance as raw OpenCL code while being more modular and concise than was feasible before.
-
-We conclude in Sections \ref{related} and 7 with related work and a discussion of the limitations of Ace at the time of writing, as well as a discussion of planned future work further validating Ace and building upon the concept of active typing.
-
-% a programming language intended to be used by professional end-users as well as researchers in areas related to high-performance computing and computational science.
-The primary contribution of this paper consists of describing several novel language mechanisms and novel combinations of existing mechanisms that Ace uses to cohesively address the design criteria above.
-We anticipate that many of these mechanisms and choices will also influence other languages in the future.
-%Following a principled design methodology, we justify each design decision by appeal to this full set of design criteria and citations to relevant literature.
-These criteria and methodology constitute another contribution of this work that we hope  will lead to fewer \emph{ad hoc} designs and serve as a foundation for evaluating competing designs in HPC and related domains in the future.
+The remainder of the paper is organized as follows: in Sec.
+\ref{usage}, we describe the basic structure and usage of Ace with an example library that internalizes and extends the OpenCL language.
+Then in Sec.
+\ref{att}, we show how this and other libraries are implemented by detailing the extension mechanisms within Ace.
+To explain and demonstrate the expressiveness of these mechanisms (in particular, active types) further, we continue in Sec.
+\ref{examples} by showing a diverse collection of abstractions drawn from different language paradigms that can be implemented as orthogonal libraries in Ace.
+We include functional datatypes, objects, macros, and typesafe format strings and regular expressions.
+In Sec.
+\ref{related}, we compare Ace to related work on language extensibility and metaprogramming.
+We conclude in Sec.
+\ref{discussion} by summarizing our contributions, discussing the essential features needed by a host language to support these mechanisms, and discuss limitations and potential future work.
 
 \section{Language Design and Usage}\label{usage}
+Listing \ref{map} shows an example of an Ace file.
+As promised, the top level of an Ace file is written directly in Python, requiring no modifications to the language (versions 2.6+ or 3.3+) nor features specific to CPython (so Ace supports alternative implementations like Jython and IronPython).
+This choice pays immediate dividends on line 1: Ace's package system is Python's package system, so Python's build tools (e.g.
+\verb|pip|) and package repostories (e.g.
+\verb|PyPI|) are directly available for distributing Ace libraries.
+
+The top-level statements in an Ace file, like the \verb|print| statement on line 10, are executed to control the compile-time behavior, rather than the run-time behavior, of the program.
+That is, Python serves as the \emph{compile-time metalanguage} (and, as we will see shortly, the \emph{type-level language}) of Ace.
+%For readers familiar with C/C++, Python can be thought of as serving a role similar to (but more general than) its preprocessor and template system (as we will see).
+Functions containing run-time behavior, like \verb|map|, are governed by a semantics that can differ from Python's (in ways that we will describe below), but they share Python's syntax.
+As a consequence, users of Ace  immediately benefit from an ecosystem of well-developed tools that work with Python code, including parsers, code highlighters, editor modes, style checkers and documentation generators.
+
+\subsection{OpenCL as an Active Library}
+The code in this section uses \verb|clx|, an example library implementing the semantics of the OpenCL programming language and extending it with some additional useful types, which we will discuss shortly.
+Ace itself provides no built-in support for OpenCL.
+
+To briefly review, in OpenCL, developers can define functions, called {\em kernels}, that execute on \emph{devices} like GPUs or multi-core CPUs \cite{opencl11}.
+Code on the host CPU launches a kernel over tens of thousands of threads.
+Each thread executes the same kernel but has access to a unique index, called its \emph{global id} (a data-parallel SPMD programming model).
+Kernel code is written in a variant of C99 extended with some new primitive types and operators, which we will introduce as needed in our examples below.
+
+\subsection{Generic Functions}
 \begin{codelisting}
-\lstinputlisting{hello.py}
-\caption{\texttt{[hello.py]} A basic Ace program demonstrating the two-phase structure of Ace programs and libraries.}
-\label{hello}
+\lstinputlisting[linebackgroundcolor={\btLstHL{5-8}}]{listing3.py}
+\caption{[\texttt{listing\ref{map}.py}] A generic data-parallel higher-order map function written using the OpenCL user module.}
+\label{map}
 \end{codelisting}
-\begin{codelisting}
-\begin{lstlisting}[style=Bash]
-$ `acec hello.py`
-Hello, compile-time world!
-Goodbye, compile-time world!
-$ `cat hello.cl`
-__kernel void main() {
-    char* hello = "Hello, run-time world!";
-    printf(hello);
-}\end{lstlisting}
-\caption{Compiling \texttt{hello.py} using the \texttt{acec} compiler.}
-\label{helloout}
-\end{codelisting}
-A variant of the standard ``Hello, World!'' example written in Ace is shown in Listing \ref{hello} and its compilation to statically-typed OpenCL kernel code is demonstrated in Listing \ref{helloout}.
+Lines 3 and 4 introduce the definition of  \verb|map|, an Ace function taking three arguments, \verb|input|, \verb|output| and \verb|f|, governed by the active base \verb|clx.std_base|.
+The body of this function, highlighted in grey for emphasis, does not have Python's semantics.
+Instead, it will be governed by the active base as well as logic associated with the types of the arguments and expressions it contains.
+
+Generic functions in Ace are instances of \verb|ace.GenericFn|.
+Generic functions consist of an abstract syntax tree, an active base and an active target\todo{call it this in intro}.
+The job of the \emph{decorator} on line 3 is to replace the Python function on lines 4-8 with an Ace generic function having the same syntax tree and the specified active base and active target.
+Because no active target was provided, the active base  provides a default target in this case.
+The default target provided by \verb|clx.std_base| is \verb|clx.opencl|, but, as we will discuss in Sec.
+\ref{bases}, other bases, like \verb|clx.cuda|, a vendor-specific language similar to OpenCL, as well as Python itself (\verb|ace.python|), can also be defined in libraries.
+One base can support multiple targets.
+
+Python decorators are simply syntactic sugar for applying a function (here, the result of \verb|ace.fn(clx.std_base)|) directly to the function definition being decorated \cite{python-decorators}.
+In other words, line 3 could be removed and replaced by inserting the equivalent statement \verb|map = ace.fn(clx.std_base)(map)| to line 9.
+ The decorator applied here extracts the abstract syntax tree from \verb|map| using the Python standard  library packages  \verb|inspect| (to retrieve source code) and \verb|ast| (to parse it into a syntax tree).
+
+\subsection{Concrete Functions}
+The definition of \verb|map| in Listing \ref{map} could be meaningful for many different type assignments to \verb|input|, \verb|output| and \verb|f|.
+The body of the function imposes certain constraints on these types, but there is not necessarily a unique type assignment that satisfies them.
+In a sense, then, \verb|map| is a \emph{family} of functions.
+Any type assignment for which the operations in the function's body are well-defined can be generated from \verb|map|.
+We call an Ace function for which types have been assigned a \emph{concrete function}.
+Concrete functions are instances of \verb|ace.ConcreteFn|.
+We will show different ways types can be assigned to the arguments of a generic function, and how this, in tandem with the active base, controls the semantics and compilation of the function body in Secs.
+\ref{procedural} and \ref{implicit}.
+First, we must understand how types are defined and used in Ace.
+
+\subsection{Types}
+In Ace, types are themselves values that can be manipulated at compile-time (unlike most contemporary languages, where user-defined types are simply declared at compile-time and cannot be defined or manipulated programmatically).
+In particular, types are instances of a Python class that implements the \verb|ace.ActiveType| interface.
+Python, then, serves also as the \emph{type-level language} of Ace (see related work in Sec.
+\ref{related}).
+
+Listing 2 shows three types, on lines 7-9, being assigned to variables for convenience (this is similar to \verb|typedef| in C-like languages or \verb|type| in Haskell).
+These types all come from the \verb|clx| library imported on line 1, none are built into Ace itself.
+
+The type named \verb|F| on line 7 corresponds to a type in OpenCL, \verb|global float*|: a pointer to a 32-bit floating point number stored in the compute device's global memory (one of four memory spaces defined by OpenCL, see \cite{opencl11}).
+Here, \verb|clx.float| is an Ace type corresponding to \verb|float| in OpenCL, and \verb|clx.Ptr| is an Ace type family corresponding to pointer types, indexed by both an address space (here, \verb|clx.global_|) and the type being pointed to.
+The \verb|clx| library contains a complete implementation of the OpenCL type system, which is based on C99.
+We will return to how this works in Section \ref{att}.
+We note that we are being more verbose than necessary for pedagogical reasons.
+The \verb|clx| library defines more concise naming schemas for OpenCL's built-in types: \verb|F| could be written \verb|clx.gp(clx.f32)|.
+We also note that Ace is \emph{unopinionated} about issues like memory safety.
+We will show how to implement safer alternatives to raw pointers in Sec.
+\ref{examples}, but we allow users to choose a low level of abstraction when they wish.(see Sec.
+\ref{discussion}).
+
+Listing \ref{mapadd5dbl} shows how to assign types to the arguments of \verb|map| so that this logic can be typechecked and compiled.
+
+This notion of types as metalanguage objects is key to the Ace compilation model and also enables other mechanisms that we will discuss in subsequent sections.
+
+Running \verb|acec listing3.py| would produce no output.
+To create a {\em concrete function} (that is, an instance of the class \verb|ace.ConcreteFn|) that can be emitted by the compiler, types must be assigned to each of the arguments.
+Listing \ref{mapadd5dbl} shows how to use the \verb|compile| method to specialize \verb|map| in \emph{two} different ways to apply the \verb|add5| function, defined on Lines 4-6, to arrays  that reside in global memory (OpenCL associates a memory space with pointer types).
+Line 9 results in a version specialized for arrays of \verb|double|s and Line 10 results in a version for arrays of \verb|int|s.
+The output of compilation is shown in Listing \ref{mapout}.
+
+The \verb|map| function defined in Listing \ref{map} will compile to one or more kernels that applies a transfer function, \verb|f|, to the element of the input array, \verb|input|, corresponding to its global id.
+It writes the result of this call into the corresponding location in the provided output array, \verb|output|.
+If the kernel is running in thread 0, it prints a message.
+The functions \verb|get_global_id| and \verb|printf| are OpenCL primitive functions.
+To emphasize again that the semantics of this code differs from Python's, we note that these functions have no meaning in Python.
+
+%\begin{codelisting}
+%\lstinputlisting{hello.py}
+%\caption{\texttt{[hello.py]} A basic Ace program demonstrating the two-phase structure of Ace programs and libraries.}
+%\label{hello}
+%\end{codelisting}
+%A variant of the standard ``Hello, World!'' example written in Ace is shown in Listing \ref{hello} and its compilation to statically-typed OpenCL kernel code is demonstrated in Listing \ref{helloout}.
 The OpenCL kernel language is a variant of C99 with some additions and restrictions to facilitate execution on GPUs and other accelerators, in addition to conventional CPUs.
 We will discuss it further in subsequent examples.
-We emphasize that this module, which we use throughout the paper, is simply a library like any other.
+%We emphasize that this module, which we use throughout the paper, is simply a library like any other.
 The core of Ace gives no special treatment to it; it is distributed together with Ace for convenience.
@@
 
-This example demonstrates several key design decisions that characterize Ace: static typing of run-time behavior, a Python-based syntax, a phase distinction between compile-time and run-time logic, and programmatic compilation.
-We discuss each of these in the next three sections.
-\subsection{Static Typing and Syntax}
-Static type systems are powerful tools for programming language design and implementation.
-By tracking the type of a value statically, a typechecker can verify the absence of many kinds of errors over all inputs.
-This simplifies and increases the performance of the run-time system, as errors need not be detected dynamically using tag checks and other kinds of assertions.
-Many parallel programming abstractions are defined in terms of, or benefit from, a type system that enforces a communication protocol, ensures the consistency of data and simplifies the dynamics of the run-time system (see Section \ref{usecases} for examples).
-Because \textbf{verifiability} and \textbf{performance} are key criteria and static typing is a core technique, Ace is fundamentally statically-typed.
+This example demonstrates several key design decisions that characterize Ace: static typing of run-time behavior, a Python-based syntax, a phase distinction between compile-time and run-time logic, and programmatic
+\subsection{Programmatic Compilation}
+\begin{codelisting}
+\lstinputlisting[linebackgroundcolor={\btLstHL{5-5}}]{listing4.py}
+\caption{[\texttt{listing\ref{mapadd5dbl}.py}] The generic \texttt{map} function compiled to map the \texttt{add\_5} function over two  types of input.}
+\label{mapadd5dbl}
+\end{codelisting}
 
-It is legitimate to ask, however, why dynamically-typed languages are so widely-used in HPC.
-Although slow and difficult to reason about, these languages generally excel at satisfying the criteria of \textbf{ease-of-use}.
-More specifically, Cordy identified the principle of \emph{conciseness} as elimination of
-redundancy and the availability of reasonable defaults \cite{cordy1992hints}.
-Statically-typed languages, particularly those that HPC programmers are exposed to, are verbose, requiring explicit and often redundant type annotations on each function and variable declaration, separate header files, explicit template headers and  instantiation and other sorts of annotations.
-The dynamically-typed languages used in HPC, on the other hand, avoid most of this overhead by relying on support from the run-time system.
-Ace was first conceived to explore the question: \emph{does conciseness require run-time mechanisms, or can one develop a statically-typed language with the same low-level memory and execution model of C but syntactic overhead comparable to a high-level scripting language?
-}
-
-Rather than designing a new syntax, or modifying the syntax of C, we chose to utilize, \emph{without modification}, the syntax of an existing language, Python.
-This choice was not arbitrary, but rather a key means by which Ace achieves both \textbf{ease-of-use} and \textbf{continuity}.
-Python's whitespace-delimited syntax is widely regarded as both concise and readable, and Python is amongst the most widely-adopted languages in computational science \cite{oliphant2007python}.
-By directly adopting Python's syntax, Ace's syntax is immediately \emph{familiar} and \emph{acceptable} to a significant segment of the intended audience.
-Moreover, a key benefit of adopting it without modifications is that any {tools} that handle Python source code, including parsers, editors, style checkers and documentation generators, can be used on Ace code without modification.
-
-\subsection{Phase Separation}
 \begin{codelisting}
-\lstinputlisting{listing3.py}
-\caption{[\texttt{listing\ref{map}.py}] A generic data-parallel higher-order map function written using the OpenCL user module.}
-\label{map}
-\end{codelisting}
-\begin{codelisting}
-\lstinputlisting{listing4.py}
-\caption{[\texttt{listing\ref{mapadd5dbl}.py}] The generic \texttt{map} function compiled to map the \texttt{add5} function over two  types of input.}
-\label{mapadd5dbl}
+\begin{lstlisting}[style=Bash]
+$ `acec listing2.py`
+Hello, compile-time world!
+[acec] listing2.cl successfully generated.
+\end{lstlisting}
+\caption{Compiling \texttt{listing2.py} using the \texttt{acec} compiler.}
+\label{mapc}
 \end{codelisting}
@@
 \lstinputlisting[style=OpenCL]{listing5.cl}
-\caption{[\texttt{listing\ref{mapadd5dbl}.cl}] The OpenCL code generated by running \texttt{acec listing\ref{mapadd5dbl}.py}.}
+\caption{[\texttt{listing\ref{mapadd5dbl}.cl}] The OpenCL file generated by Listing \ref{mapc}.}
 \label{mapout}
 \end{codelisting}
-
-%To demonstrate the somewhat unconventional structure of Ace programs and libraries, we begin in Listing \ref{hello} with a variant of the canonical ``Hello, World!'' example written using the \verb|OpenCL| module.
-
-Ace's \textbf{continuity} with Python does not stop at its syntax. 
-Perhaps the most immediately apparent departure from the standard ``Hello, World!'' example comes on lines 3 and 10 of Listing \ref{hello}, which contain Python \verb|print| statements that are executed at {\em compile-time}.
-Ace programs and libraries are Python scripts at the top-level, rather than a list of declarations as in most conventional languages.
-In other words, Python is the {\em compile-time metalanguage} of Ace.
-
-A consequence of this choice is that Ace can leverage Python's well-developed package system and associated distribution infrastructure directly (e.g.
-Line 1).
-This serves to address another key \textbf{ease-of-use} issue often associated with C-based languages: the fragility of the preprocessor-based packaging system they historically have relied upon.
-
-Ace functions defining run-time logic are introduced by a decorator.
-The \verb|@OpenCL.fn| decorator on Line 5 indicates that the \verb|main| function is a statically-typed Ace function targeting the OpenCL backend for code generation (Section \ref{backend}).
-Without this decorator, the function would simply be a conventional Python function that could be called only at compile-time.
-Doing so would fail here: \verb|printf| is not a Python function.
-Rather, it is a compile-time Python object representing an OpenCL primitive residing in the \verb|OpenCL| module.
-It has an associated Ace type that controls what types of input it can receive and the type of its output given its input type (see Section \ref{att}).
-
-\subsection{Programmatic Compilation}
 As defined on Lines 5-7, \verb|main| is a \emph{generic function}.
@@
 
-\subsection{Example 2: Higher-Order Map}
-The ``Hello, World!'' example demonstrates the structure of Ace programs, but it does not require working with types.
-Listing \ref{map} shows an imperative, data-parallel map primitive written using the OpenCL library introduced above.
-To review, in OpenCL users can define functions, called {\em kernels}, that execute across thousands of threads.
-Each kernel has access to a unique index, called its {\em global id}, which can be to ensure that each thread operates on different parts of the input data (Line 5).
-The \verb|map| kernel defined in Listing \ref{map} applies a transfer function, \verb|f|, to the element of the input array, \verb|input|, corresponding to its global id.
-It writes the result of this call into the corresponding location in the provided output array, \verb|output|.
-
-As above, \verb|map| is a {\em generic function} (specifically, it is an instance of the class \verb|ace.GenericFn|).
-This means that its arguments have not been assigned types.
-The functionality given by the \verb|map| definition is in fact applicable to many combinations of types for \verb|input| and \verb|output| and functions, \verb|f|.
-In this sense, \verb|map| is actually a \emph{family} of functions defined for all types assignments for \verb|input|, \verb|output| and \verb|f| such that the operations in the function's body are well-defined.
-
-Running \verb|acec listing3.py| would produce no output.
-To create a {\em concrete function} (that is, an instance of the class \verb|ace.ConcreteFn|) that can be emitted by the compiler, types must be assigned to each of the arguments.
-Listing \ref{mapadd5dbl} shows how to use the \verb|compile| method to specialize \verb|map| in \emph{two} different ways to apply the \verb|add5| function, defined on Lines 4-6, to arrays  that reside in global memory (OpenCL associates a memory space with pointer types).
-Line 9 results in a version specialized for arrays of \verb|double|s and Line 10 results in a version for arrays of \verb|int|s.
-The output of compilation is shown in Listing \ref{mapout}.
-
-\subsection{Types as Metalanguage Objects}
-The \verb|compile| method assigns types to the arguments of a generic function.
-In Listing \ref{mapadd5dbl}, the types we are using are given shorter names, for convenience, on Line 8 (that is, variables in the metalanguage can be used like \verb|typedef| is used in a C-like language).
-The types \verb|int| and \verb|double| imported from the \verb|OpenCL| module correspond to the OpenCL types of the same name.
-The types \verb|gptr(int)| and \verb|gptr(double)| correspond to \verb|__global int*| and \verb|__global double*|.
-
 These types are \emph{objects in the metalanguage}, Python.
@@
 The types \verb|double| and \verb|int| are instances of \verb|OpenCL.FloatType| and \verb|OpenCL.IntegerType|, respectively.
-This notion of types as metalanguage objects is key to the Ace compilation model and also enables other mechanisms that we will discuss in subsequent sections.
-
 \subsection{Type Propagation}
@@
 \subsection{Whole-Function Type Inference}
+\begin{codelisting}[t]
+\lstinputlisting[commentstyle=\color{mauve}]{listing7.py}
+\caption{[\texttt{listing\ref{metaprogramming}.py}] Metaprogramming with Ace, showing how to construct generic functions from both strings and abstract syntax trees, and how to manipulate syntax trees at compile-time.}
+\label{metaprogramming}
+\end{codelisting}
 \begin{codelisting}
+\lstinputlisting{listing8.py}
+\caption{[\texttt{listing\ref{py}.py}] A full OpenCL program using the \texttt{Ace.OpenCL} Python bindings, including data transfer to and from a device and direct invocation of a generic function, \texttt{map}, as a kernel without explicit compilation.}
+\label{py}
+\end{codelisting}
+
+%\begin{codelisting}
 \lstinputlisting{listing6.py}
-\caption{\texttt{[listing6.py]} A function demonstrating whole-function type inference when multiple values with differing types are assigned to a single identifier, \texttt{y}.}
-\label{inference}
-\end{codelisting}
+%\caption{\texttt{[listing6.py]} A function demonstrating whole-function type inference when multiple values with differing types are assigned to a single identifier, \texttt{y}.}
+%\label{inference}
+%\end{codelisting}
 
@@
 Note that this example would also work correctly if the assignments to \verb|y| were replaced with \verb|return| statements (in other words, the return value of a function is treated as an assignable for the purpose of type inference).
+
+\subsection{Implicit Compilation and Execution}\label{compenv}\label{backend}
+Professional end-users today generally use dynamically-typed high-level languages like MATLAB, Python, R or Perl for tasks that are not performance-sensitive, such as small-scale data analysis and plotting \cite{nguyen2010survey}.
+For portions of their analyses where the performance overhead of dynamic type checking and automatic memory management is too high, they will typically call into code written in a statically-typed, low-level language, most commonly C or Fortran, that uses low-level parallel abstractions like pthreads and MPI \cite{4222616,basili2008understanding}.
+Unfortunately, these low-level languages and abstractions are notoriously difficult to use and automatic verification is intractable in general.\todo{integrate this}
+
+As discussed in the Introduction, a common workflow for professional end-users is to use a high-level scripting language for orchestration, small-scale data analysis and visualization and call into a low-level language for performance-critical sections.
+Python is designed for this style of use \cite{sanner1999python} and is widely used by professional end-users in HPC as a high-level scripting language.
+It features mature support for calling into code written in low-level languages.
+Developers can call into native libraries using its foreign function interface (FFI),  or by using a wrapper library like \verb|pycuda| for code compiled with CUDA \cite{klockner2011pycuda} or \verb|weave| for C and C++.
+
+Although C, C++ and CUDA's compilers are separate executables on the system, the OpenCL language was also designed for this workflow, in that it exposes the compiler and memory management directly as an API, called the \emph{host API}.
+The \verb|pyopencl| module exposes this API and supports basic interoperability with \verb|numpy|, the low-level linear algebra package for Python.
+With both \verb|pyopencl| and \verb|pycuda|, users generate OpenCL or CUDA source code as strings, compile it programmatically, then execute it using the run-time APIs that each library provides \cite{klockner2011pycuda}.
+This mode of use can be considered one where Python serves as an \emph{interactive compilation environment}.
+
+Ace supports a refinement to this workflow, as an alternative to the \verb|acec| compiler described above.
+In other words, \verb|acec| can generate source code for kernels but it does not specify how and from what language thet will be called.
+This mode of use allows Python to be the calling language.
+
+At the time of writing, this is only supported with the OpenCL backend, but other backends can implement this feature as well by satisfying a simple interface specification.
+The OpenCL host API wraps and is a superset of the \verb|pyopencl|, adding a simpler type-aware API.
+Both generic functions and concrete functions can then be called like regular Python functions, with additional keyword arguments specifying the global and local size (i.e.
+the number of threads and how they are grouped).
+Device buffers can carry type information, unlike in the basic OpenCL host API, and this type information can be propagated from Python to Ace functions directly, as if the call had been within a kernel,  eliminating the \verb|compile| step that we have been using thus far and thus all mention of types.
+
+An example of this for the generic \verb|map| function defined in Listing \ref{map} is shown in Listing \ref{py}, with the call itself on Lines 12-13. 
+The first two arguments to \verb|map| are OpenCL buffers, generated using a simplified wrapper to the \verb|pyopencl| APIs on Lines 9-10.
+This wrapper associates type information with each buffer, based on the type of the \verb|numpy| array, and this is used to implicitly compile \verb|map| as appropriate the first time it is called for any given combination of input types.
+Notice also that \verb|add5| is passed in directly.
+
+By way of comparison, the same program written using the OpenCL C API directly is an order of magnitude larger and significantly more difficult to read and understand.
+It does not support higher-order functions nor is there any way to write \verb|map| in a type-generic way.
+A full implementation of the logic of \verb|map| written using the \verb|pyopencl| bindings and metaprogramming techniques as described in \cite{klockner2011pycuda} is still twice as large and more difficult to comprehend than the code we have shown thus far.
+Not shown are several additional conveniences, such as delegated kernel sizing and \verb|In| and \verb|Out| constructs that can reduce the size and improve the clarity of this code further; due to a lack of space, the reader is referred to the language documentation.
+
+\subsection{Metaprogramming in Ace}
+Metaprogramming refers to the practice of writing programs that manipulate other programs.
+There are a number of use cases for this technique, including domain-specific optimizations and code generation for programs with a repetitive structure that cannot easily be captured using available abstractions \cite{klockner2011pycuda}.
+OpenCL in particular relies on code generation as a fundamental mechanism, which is cited as justification for its lack of support for higher-order programming.
+Ace supports programmatic compilation, higher-order functions and a flexible language extension mechanism, so many use cases for metaprogramming have been eliminated.
+
+However, there are still valid uses for this feature and Ace fully supports it.
+On Lines 4-7 of Listing \ref{metaprogramming}, an Ace function is constructed from a string containing its source using the \verb|from_str| variant of the \verb|OpenCL.fn| method.
+It can also be constructed directly from an abstract syntax tree (AST), as implemented by the Python standard \verb|ast| package, using the \verb|from_ast| variant of \verb|fn|, demonstrated on Line 10.
+The AST here is generated programmatically by calling the \verb|specialize| function, which produces a copy of the syntax tree of the \verb|plus| function with the argument, \verb|b|, eliminated and its uses replaced with a constant, \verb|5|.
+This transformation as well as some others are distributed in \verb|ace.astx| for convenience.
 
@@
 
-\subsection{Use Cases}\label{usecases}
+\section{Examples}\label{examples}
 The development of the full OpenCL language using only the extension mechanisms described above provides evidence of the power of this approach.
@@
 
-\section{Ace: a Compilation Environment}\label{compenv}\label{backend}
-\begin{codelisting}
-\lstinputlisting{listing8.py}
-\caption{[\texttt{listing\ref{py}.py}] A full OpenCL program using the \texttt{Ace.OpenCL} Python bindings, including data transfer to and from a device and direct invocation of a generic function, \texttt{map}, as a kernel without explicit compilation.}
-\label{py}
-\end{codelisting}
-
-As discussed in the Introduction, a common workflow for professional end-users is to use a high-level scripting language for orchestration, small-scale data analysis and visualization and call into a low-level language for performance-critical sections.
-Python is designed for this style of use \cite{sanner1999python} and is widely used by professional end-users in HPC as a high-level scripting language.
-It features mature support for calling into code written in low-level languages.
-Developers can call into native libraries using its foreign function interface (FFI),  or by using a wrapper library like \verb|pycuda| for code compiled with CUDA \cite{klockner2011pycuda} or \verb|weave| for C and C++.
-
-Although C, C++ and CUDA's compilers are separate executables on the system, the OpenCL language was also designed for this workflow, in that it exposes the compiler and memory management directly as an API, called the \emph{host API}.
-The \verb|pyopencl| module exposes this API and supports basic interoperability with \verb|numpy|, the low-level linear algebra package for Python.
-With both \verb|pyopencl| and \verb|pycuda|, users generate OpenCL or CUDA source code as strings, compile it programmatically, then execute it using the run-time APIs that each library provides \cite{klockner2011pycuda}.
-This mode of use can be considered one where Python serves as an \emph{interactive compilation environment}.
-
-Ace supports a refinement to this workflow, as an alternative to the \verb|acec| compiler described above.
-In other words, \verb|acec| can generate source code for kernels but it does not specify how and from what language thet will be called.
-This mode of use allows Python to be the calling language.
-
-At the time of writing, this is only supported with the OpenCL backend, but other backends can implement this feature as well by satisfying a simple interface specification.
-The OpenCL host API wraps and is a superset of the \verb|pyopencl|, adding a simpler type-aware API.
-Both generic functions and concrete functions can then be called like regular Python functions, with additional keyword arguments specifying the global and local size (i.e.
-the number of threads and how they are grouped).
-Device buffers can carry type information, unlike in the basic OpenCL host API, and this type information can be propagated from Python to Ace functions directly, as if the call had been within a kernel,  eliminating the \verb|compile| step that we have been using thus far and thus all mention of types.
-
-An example of this for the generic \verb|map| function defined in Listing \ref{map} is shown in Listing \ref{py}, with the call itself on Lines 12-13. 
-The first two arguments to \verb|map| are OpenCL buffers, generated using a simplified wrapper to the \verb|pyopencl| APIs on Lines 9-10.
-This wrapper associates type information with each buffer, based on the type of the \verb|numpy| array, and this is used to implicitly compile \verb|map| as appropriate the first time it is called for any given combination of input types.
-Notice also that \verb|add5| is passed in directly.
-
-By way of comparison, the same program written using the OpenCL C API directly is an order of magnitude larger and significantly more difficult to read and understand.
-It does not support higher-order functions nor is there any way to write \verb|map| in a type-generic way.
-A full implementation of the logic of \verb|map| written using the \verb|pyopencl| bindings and metaprogramming techniques as described in \cite{klockner2011pycuda} is still twice as large and more difficult to comprehend than the code we have shown thus far.
-Not shown are several additional conveniences, such as delegated kernel sizing and \verb|In| and \verb|Out| constructs that can reduce the size and improve the clarity of this code further; due to a lack of space, the reader is referred to the language documentation.
-
-\subsection{Metaprogramming in Ace}
-\begin{codelisting}
-\lstinputlisting[commentstyle=\color{mauve}]{listing7.py}
-\caption{[\texttt{listing9.py}] Metaprogramming with Ace, showing how to construct generic functions from both strings and abstract syntax trees, and how to manipulate syntax trees at compile-time.}
-\label{metaprogramming}
-\end{codelisting}
-Metaprogramming refers to the practice of writing programs that manipulate other programs.
-There are a number of use cases for this technique, including domain-specific optimizations and code generation for programs with a repetitive structure that cannot easily be captured using available abstractions \cite{klockner2011pycuda}.
-OpenCL in particular relies on code generation as a fundamental mechanism, which is cited as justification for its lack of support for higher-order programming.
-Ace supports programmatic compilation, higher-order functions and a flexible language extension mechanism, so many use cases for metaprogramming have been eliminated.
-
-However, there are still valid uses for this feature and Ace fully supports it.
-On Lines 4-7 of Listing \ref{metaprogramming}, an Ace function is constructed from a string containing its source using the \verb|from_str| variant of the \verb|OpenCL.fn| method.
-It can also be constructed directly from an abstract syntax tree (AST), as implemented by the Python standard \verb|ast| package, using the \verb|from_ast| variant of \verb|fn|, demonstrated on Line 10.
-The AST here is generated programmatically by calling the \verb|specialize| function, which produces a copy of the syntax tree of the \verb|plus| function with the argument, \verb|b|, eliminated and its uses replaced with a constant, \verb|5|.
-This transformation as well as some others are distributed in \verb|ace.astx| for convenience.
-
 \section{Related Work}\label{related}
+\begin{figure*}
+\begin{tabular}[t]{ l | l | c c c c c }
+\textbf{Approach} & \textbf{Examples} & Library & {Extensible} & Extensible & Composable & Alternative \\
+\hline
+Active Types & Ace & x & ~ & x & x & x\\
+Desugaring & SugarJ, Sugar* & ~ & x & ~ & ~ & ~\\
+Rule Injection & Qi, Typed Racket/Clojure, A? & 1+2 & ~ & x & ~ & ~\\
+Static metaprogramming & Scala macros, OJ, OC++ & ~ & ~ & ~ & x & ~\\
+Cross-compilation & Delite & x & ~ & ~ & ~ & x\\
+EDSL frameworks & ? & x & x & x & ~ & ~\\
+Type-specific literals & Wyvern & ~ & x & ~ & x & ~\\
+\end{tabular}
+\caption{Comparison to related approaches}
+\end{figure*}
 \subsection{Active Libraries}
@@
 
-\section{Conclusion}
+\section{Discussion}\label{discussion}
+Static type systems are powerful tools for programming language design and implementation.
+By tracking the type of a value statically, a typechecker can verify the absence of many kinds of errors over all inputs.
+This simplifies and increases the performance of the run-time system, as errors need not be detected dynamically using tag checks and other kinds of assertions.
+Many parallel programming abstractions are defined in terms of, or benefit from, a type system that enforces a communication protocol, ensures the consistency of data and simplifies the dynamics of the run-time system (see Section \ref{usecases} for examples).
+Because \textbf{verifiability} and \textbf{performance} are key criteria and static typing is a core technique, Ace is fundamentally statically-typed.
+
+It is legitimate to ask, however, why dynamically-typed languages are so widely-used in HPC.
+Although slow and difficult to reason about, these languages generally excel at satisfying the criteria of \textbf{ease-of-use}.
+More specifically, Cordy identified the principle of \emph{conciseness} as elimination of
+redundancy and the availability of reasonable defaults \cite{cordy1992hints}.
+Statically-typed languages, particularly those that HPC programmers are exposed to, are verbose, requiring explicit and often redundant type annotations on each function and variable declaration, separate header files, explicit template headers and  instantiation and other sorts of annotations.
+The dynamically-typed languages used in HPC, on the other hand, avoid most of this overhead by relying on support from the run-time system.
+Ace was first conceived to explore the question: \emph{does conciseness require run-time mechanisms, or can one develop a statically-typed language with the same low-level memory and execution model of C but syntactic overhead comparable to a high-level scripting language?
+}
+
+Rather than designing a new syntax, or modifying the syntax of C, we chose to utilize, \emph{without modification}, the syntax of an existing language, Python.
+This choice was not arbitrary, but rather a key means by which Ace achieves both \textbf{ease-of-use} and \textbf{continuity}.
+Python's whitespace-delimited syntax is widely regarded as both concise and readable, and Python is amongst the most widely-adopted languages in computational science \cite{oliphant2007python}.
+By directly adopting Python's syntax, Ace's syntax is immediately \emph{familiar} and \emph{acceptable} to a significant segment of the intended audience.
+Moreover, a key benefit of adopting it without modifications is that any {tools} that handle Python source code, including parsers, editors, style checkers and documentation generators, can be used on Ace code without modification.
+Researchers often dismiss the importance of syntax.
+By using a well-developed syntax, they no longer need to worry about the equally ``trivial'' task of implementing tools for it.
 
@@
 We hope that Ace will be developed further by the community to strengthen the foundations upon which new abstractions are implemented and deployed into the HPC professional end-user community.
+
+We suggest three mutually-related {design criteria} that, unlike those in bold above, many languages and language-integrated abstractions have failed to adequately consider: \textbf{continuity}, \textbf{extensibility} and \textbf{interoperability}.
+These criteria encompass the intuitions that new abstractions will not be adopted in a vacuum, that programming systems must  support change, and that interacting components of an application or workflow should be able to make use of different abstractions naturally and without the possibility of conflict arising at their interface boundaries.
+
+ We anticipate that coding guidelines mandating the use of abstractions that can be shown to have certain desirable properties will replace language-mandated enforcement opinions
 

### 2013-11-13 ace-pldi14/listing3.py

@@ -1,6 +1,10 @@
-from ace.OpenCL import OpenCL, get_global_id
+import ace, examples.clx as clx
 
-@OpenCL.fn
+@ace.fn(clx.std_base)
 def map(input, output, f):
-    gid = get_global_id(0)
-    output[gid] = f(input[gid])
+    thread_idx = get_global_id()
+    output[thread_idx] = f(input[thread_idx])
+    if thread_idx == 0:
+    	printf("Hello, run-time world!")
+
+print "Hello, compile-time world!"

### 2013-11-13 ace-pldi14/listing4.py

@@ -1,10 +1,12 @@
-from listing3 import map
-from ace.OpenCL import gptr, double, int
+import listing1, ace, examples.clx as clx
 
-@OpenCL.fn
-def add5(x):
+@ace.fn(clx.std_base)
+def add_5(x):
     return x + 5
 
-D = gptr(double); I = gptr(int); A = add5.ace_type
-map_add5_dbl = map.compile(D, D, A)
-map_add5_int = map.compile(I, I, A)
+F = clx.Ptr(clx.global_, clx.float)
+C = clx.Ptr(clx.global_, clx.cplx(clx.int))
+A = add_5.ace_type
+
+map_add5_float = listing1.map[[F, F, A]]
+map_add5_c_int = listing1.map[[C, C, A]]

### 2013-11-13 ace-pldi14/listing7.py

@@ -1,10 +1,6 @@
-from ace.OpenCL import OpenCL
-import ace.astx as astx
+import astx, ace, examples.clx as clx
 
-plus = OpenCL.fn.from_str("""
-def plus(a, b):
-    return a + b
-""")
-
-add5_ast = astx.specialize(plus.ast, b=5)
-add5 = OpenCL.fn.from_ast(add5_ast)
+plus = ace.fn(clx.std_base)("""def plus(a, b):
+    return a + b""")
+add_5_ast = astx.specialize(plus.ast, b=5)
+add_5 = ace.fn(clx.std_base)(add5_ast)

### 2013-11-13 ace-pldi14/listing8.py

@@ -1,16 +1,12 @@
-import numpy as np
-import ace.OpenCL.bindings as cl
-from listing3 import map
-from listing4 import add5
+import examples.clx as clx, numpy, listing1, listing5
 
-cl.ctx = cl.Context.for_device(0, 0)
+clx.default_ctx = clx.Context.for_device(0, 0)
 
-input = np.ones((1024,))
-d_in = cl.to_device(input)
-d_out = cl.alloc(like=d_in)
+input = numpy.ones((1024,))
+d_input = clx.to_device(input)
+d_output = clx.alloc(like=input)
 
-map(d_in, d_out, add5, 
-    global_size=d_in.shape, local_size=(128,))
+listing1.map(d_input, d_output, listing5.add_5, 
+  global_size=d_in.shape, local_size=(128,))
 
-out = cl.from_device(d_out)
-assert (out == input + 5).all()
+assert (cl.from_device(d_out) == input + 5).all()

### 2013-11-14 ace-pldi14/ace-pldi14.tex
happy thru sec 2.3 metaprogramming

@@
 We will show examples of Ace targeting Python as well as OpenCL and CUDA, lower-level languages often used to program graphics hardware.
-An active base or type can support multiple target languages.
+An active base or type can support multiple \emph{active targets}, which mediate translation of Ace code to code in a target language.
 Ace functions targeting a language with Python bindings can be called directly from Python scripts, with compilation occurring implicitly.
-In some cases, types can propagate from Python into Ace.
+For some data structures, types can propagate from Python into Ace.
 We show how this can be used to streamline the kinds of interactive workflows that Python is often used for.
-Ace can also be used non-interactively from the shell, producing source files that can be further compiled and executed by external tools.
+Ace can also be used non-interactively from the shell, producing source files that can be further compiled and executed by external means.
 
@@
 We conclude in Sec.
-\ref{discussion} by summarizing our contributions, discussing the essential features needed by a host language to support these mechanisms, and discuss limitations and potential future work.
+\ref{discussion} by summarizing our contributions, discussing the essential features needed by a host language to support these mechanisms, and describing their limitations and potential future work.
 
@@
 Functions containing run-time behavior, like \verb|map|, are governed by a semantics that can differ from Python's (in ways that we will describe below), but they share Python's syntax.
-As a consequence, users of Ace  immediately benefit from an ecosystem of well-developed tools that work with Python code, including parsers, code highlighters, editor modes, style checkers and documentation generators.
+As a consequence, users of Ace  immediately benefit from an ecosystem of well-developed tools that work with Python syntax, including parsers, code highlighters, editor modes, style checkers and documentation generators.
 
@@
 The code in this section uses \verb|clx|, an example library implementing the semantics of the OpenCL programming language and extending it with some additional useful types, which we will discuss shortly.
-Ace itself provides no built-in support for OpenCL.
+Ace itself has no built-in support for OpenCL.
 
-To briefly review, in OpenCL, developers can define functions, called {\em kernels}, that execute on \emph{devices} like GPUs or multi-core CPUs \cite{opencl11}.
-Code on the host CPU launches a kernel over tens of thousands of threads.
-Each thread executes the same kernel but has access to a unique index, called its \emph{global id} (a data-parallel SPMD programming model).
+To briefly review, OpenCL provides a data-parallel SPMD programming model where developers define functions, called {\em kernels}, for execution on \emph{compute devices} like GPUs or multi-core CPUs \cite{opencl11}.
+Each thread executes the same kernel but has access to a unique index, called its \emph{global ID}.
 Kernel code is written in a variant of C99 extended with some new primitive types and operators, which we will introduce  as needed in our examples below.
@@
 \lstinputlisting[linebackgroundcolor={\btLstHL{5-8}}]{listing3.py}
-\caption{[\texttt{listing\ref{map}.py}] A generic data-parallel higher-order map function written using the OpenCL user module.}
+\caption{[\texttt{listing\ref{map}.py}] A generic data-parallel higher-order map function targeting OpenCL.}
 \label{map}
 \end{codelisting}
-Lines 3 and 4 introduce the definition of  \verb|map|, an Ace function taking three arguments, \verb|input|, \verb|output| and \verb|f|, governed by the active base \verb|clx.std_base|.
+\begin{codelisting}[t]
+\lstinputlisting[commentstyle=\color{mauve}]{listing7.py}
+\caption{[\texttt{listing\ref{metaprogramming}.py}] Metaprogramming with Ace, showing how to construct generic functions from abstract syntax trees.}
+\label{metaprogramming}
+\end{codelisting}
+Lines 3-4 introduce \verb|map|, an Ace function governed by the active base \verb|clx.base| taking three arguments.
+The \emph{active target}, \verb|clx.opencl|, determines which language the function will target (here, the OpenCL kernel language) and mediates code generation.
+
 The body of this function, highlighted in grey for emphasis, does not have Python's semantics.
-Instead, it will be governed by the active base as well as logic associated with the types of the arguments and expressions it contains.
+Instead, it will be governed by the active base together with the active types used within it.
+No such types have been provided explicitly for the arguments yet, however.
+Because our type system is extensible, the code inside could be meaningful for many different assignments of types to the arguments, so \verb|map| is not yet a \emph{concrete function}.
+We call functions awaiting types \emph{generic functions}.
+%Once types have been assigned, they are called \emph{concrete functions}.
 
-Generic functions in Ace are instances of \verb|ace.GenericFn|.
-Generic functions consist of an abstract syntax tree, an active base and an active target\todo{call it this in intro}.
-The job of the \emph{decorator} on line 3 is to replace the Python function on lines 4-8 with an Ace generic function having the same syntax tree and the specified active base and active target.
-Because no active target was provided, the active base  provides a default target in this case.
-The default target provided by \verb|clx.std_base| is \verb|clx.opencl|, but, as we will discuss in Sec.
-\ref{bases}, other bases, like \verb|clx.cuda|, a vendor-specific language similar to OpenCL, as well as Python itself (\verb|ace.python|), can also be defined in libraries.
-One base can support multiple targets.
+Generic functions are represented at compile-time as instances of \verb|ace.GenericFn| and consist of an abstract syntax tree, an {active base} and an {active target}.
+The purpose of the \emph{decorator} on line 3 is to replace the Python function on lines 4-8 with an Ace generic function having the same syntax tree and the provided active base and active target.
+Decorators in Python are simply syntactic sugar for applying the decorator function directly to the function  being decorated \cite{python}.
+In other words, line 3 could be replaced by inserting the following  statement on line 9:
 
-Python decorators are simply syntactic sugar for applying a function (here, the result of \verb|ace.fn(clx.std_base)|) directly to the function definition being decorated \cite{python-decorators}.
-In other words, line 3 could be removed and replaced by inserting the equivalent statement \verb|map = ace.fn(clx.std_base)(map)| to line 9.
- The decorator applied here extracts the abstract syntax tree from \verb|map| using the Python standard  library packages  \verb|inspect| (to retrieve source code) and \verb|ast| (to parse it into a syntax tree).
+\verb|map = ace.fn(clx.base, clx.opencl)(map)|
 
-\subsection{Concrete Functions}
-The definition of \verb|map| in Listing \ref{map} could be meaningful for many different type assignments to \verb|input|, \verb|output| and \verb|f|.
-The body of the function imposes certain constraints on these types, but there is not necessarily a unique type assignment that satisfies them.
-In a sense, then, \verb|map| is a \emph{family} of functions.
-Any type assignment for which the operations in the function's body are well-defined can be generated from \verb|map|.
-We call an Ace function for which types have been assigned a \emph{concrete function}.
-Concrete functions are instances of \verb|ace.ConcreteFn|.
-We will show different ways types can be assigned to the arguments of a generic function, and how this, in tandem with the active base, controls the semantics and compilation of the function body in Secs.
-\ref{procedural} and \ref{implicit}.
-First, we must understand how types are defined and used in Ace.
+The abstract syntax tree for \verb|map| is extracted using the Python standard  library packages  \verb|inspect| (to retrieve its source code) and \verb|ast| (to parse it into a syntax tree).
+
+\subsection{Metaprogramming in Ace}
+Generic functions can be generated directly from ASTs as well, providing Ace with support for  straightforward metaprogramming.
+Listing \ref{metaprogramming} shows how to generate two more generic functions, \verb|thresh_scale| and \verb|thresh_dbl|.
+The latter is derived from the former using a library for manipulating Python syntax trees, \verb|astx|.
+In particular, its \verb|specialize| function replaces uses of the second argument of \verb|thresh_scale| with the literal \verb|2.0| (and changes the function's name), leaving a function of one argument.
 
 \subsection{Types}
+To go from a generic function to a \emph{concrete function}, we must provide a type assignment for its arguments.
+We will show different ways to do this in Secs.
+\ref{procedural} and \ref{implicit}.
+Before we can do this, however, we must understand what types are.
+
 In Ace, types are themselves values that can be manipulated at compile-time (unlike most contemporary languages, where user-defined types are simply declared at compile-time and cannot be defined or manipulated programmatically).
@@
 
-Listing 2 shows three types, on lines 7-9, being assigned to variables for convenience (this is similar to \verb|typedef| in C-like languages or \verb|type| in Haskell).
+Listing 2 shows three types, on lines 7-9, being assigned to variables for convenience (a construct similar to \verb|typedef| in C-like languages or \verb|type| in Haskell).
 These types all come from the \verb|clx| library imported on line 1, none are built into Ace itself.
@@
 This example demonstrates several key design decisions that characterize Ace: static typing of run-time behavior, a Python-based syntax, a phase distinction between compile-time and run-time logic, and programmatic
-\subsection{Programmatic Compilation}
+\subsection{Programmatic Compilation}\label{procedural}
 \begin{codelisting}
@@
 \begin{lstlisting}[style=Bash]
-$ `acec listing2.py`
+$ `acec listing3.py`
 Hello, compile-time world!
-[acec] listing2.cl successfully generated.
+[acec] listing3.cl successfully generated.
 \end{lstlisting}
-\caption{Compiling \texttt{listing2.py} using the \texttt{acec} compiler.}
+\caption{Compiling \texttt{listing\ref{mapadd5dbl}.py} using the \texttt{acec} compiler.}
 \label{mapc}
@@
 \subsection{Whole-Function Type Inference}
-\begin{codelisting}[t]
-\lstinputlisting[commentstyle=\color{mauve}]{listing7.py}
-\caption{[\texttt{listing\ref{metaprogramming}.py}] Metaprogramming with Ace, showing how to construct generic functions from both strings and abstract syntax trees, and how to manipulate syntax trees at compile-time.}
-\label{metaprogramming}
-\end{codelisting}
 \begin{codelisting}
@@
 %\begin{codelisting}
-\lstinputlisting{listing6.py}
+%\lstinputlisting{listing6.py}
 %\caption{\texttt{[listing6.py]} A function demonstrating whole-function type inference when multiple values with differing types are assigned to a single identifier, \texttt{y}.}
@@
 
-\subsection{Implicit Compilation and Execution}\label{compenv}\label{backend}
+\subsection{Implicit Compilation and Execution}\label{compenv}\label{backend}\label{implicit}
 Professional end-users today generally use dynamically-typed high-level languages like MATLAB, Python, R or Perl for tasks that are not performance-sensitive, such as small-scale data analysis and plotting \cite{nguyen2010survey}.
@@
 
-\subsection{Metaprogramming in Ace}
-Metaprogramming refers to the practice of writing programs that manipulate other programs.
-There are a number of use cases for this technique, including domain-specific optimizations and code generation for programs with a repetitive structure that cannot easily be captured using available abstractions \cite{klockner2011pycuda}.
-OpenCL in particular relies on code generation as a fundamental mechanism, which is cited as justification for its lack of support for higher-order programming.
-Ace supports programmatic compilation, higher-order functions and a flexible language extension mechanism, so many use cases for metaprogramming have been eliminated.
+\section{Extensibility}\label{att}
+If no active target is explicitly provided, as in our example, the active base  provides a default target.
+The default target provided by \verb|clx.base| is \verb|clx.opencl|, but other targets, like \verb|clx.cuda|, a vendor-specific language similar to OpenCL, as well as Python itself (\verb|ace.python|), can also be defined in libraries.
+One base can support multiple targets.
+We will discuss targets in Sec.
+\ref{targets}.
 
-However, there are still valid uses for this feature and Ace fully supports it.
-On Lines 4-7 of Listing \ref{metaprogramming}, an Ace function is constructed from a string containing its source using the \verb|from_str| variant of the \verb|OpenCL.fn| method.
-It can also be constructed directly from an abstract syntax tree (AST), as implemented by the Python standard \verb|ast| package, using the \verb|from_ast| variant of \verb|fn|, demonstrated on Line 10.
-The AST here is generated programmatically by calling the \verb|specialize| function, which produces a copy of the syntax tree of the \verb|plus| function with the argument, \verb|b|, eliminated and its uses replaced with a constant, \verb|5|.
-This transformation as well as some others are distributed in \verb|ace.astx| for convenience.
-
-\section{Extensibility}\label{att}
 Thus far, we have been discussing the \verb|OpenCL| module in our examples.

### 2013-11-14 ace-pldi14/listing3.py

@@ -1,6 +1,6 @@
 import ace, examples.clx as clx
 
-@ace.fn(clx.std_base)
+@ace.fn(clx.base, clx.opencl)
 def map(input, output, f):
     thread_idx = get_global_id()
     output[thread_idx] = f(input[thread_idx])

### 2013-11-14 ace-pldi14/listing4.py

@@ -1,12 +1,8 @@
-import listing1, ace, examples.clx as clx
+import listing1, listing2, examples.clx as clx
 
-@ace.fn(clx.std_base)
-def add_5(x):
-    return x + 5
+T1 = clx.Ptr(clx.global_, clx.float)
+T2 = clx.Ptr(clx.global_, clx.cplx(clx.int))
+TF = listing2.thresh_dbl.ace_type
 
-F = clx.Ptr(clx.global_, clx.float)
-C = clx.Ptr(clx.global_, clx.cplx(clx.int))
-A = add_5.ace_type
-
-map_add5_float = listing1.map[[F, F, A]]
-map_add5_c_int = listing1.map[[C, C, A]]
+map_add5_float = listing1.map[[T1, T1, TF]]
+map_add5_c_int = listing1.map[[T2, T2, TF]]

### 2013-11-14 ace-pldi14/listing7.py

@@ -1,6 +1,9 @@
-import astx, ace, examples.clx as clx
+import ace, examples.clx as clx, ast, astx
 
-plus = ace.fn(clx.std_base)("""def plus(a, b):
-    return a + b""")
-add_5_ast = astx.specialize(plus.ast, b=5)
-add_5 = ace.fn(clx.std_base)(add5_ast)
+cl_fn = ace.fn(clx.base, clx.opencl)
+thresh_scale = cl_fn(ast.parse("""
+def thresh_scale(x, s):
+	return x * s if x >= 0 else 0"""))
+
+thresh_dbl = cl_fn(astx.specialize(thresh_scale.ast, 
+	"thresh_dbl", s=ast.parse("2.0"))

### 2013-11-14 ace-pldi14/listing8.py

@@ -1,4 +1,4 @@
-import examples.clx as clx, numpy, listing1, listing5
+import listing1, listing2, examples.clx as clx, numpy
 
 clx.default_ctx = clx.Context.for_device(0, 0)
 
@@ -6,7 +6,7 @@ input = numpy.ones((1024,))
 d_input = clx.to_device(input)
 d_output = clx.alloc(like=input)
 
-listing1.map(d_input, d_output, listing5.add_5, 
+listing1.map(d_input, d_output, listing2.thresh_dbl, 
   global_size=d_in.shape, local_size=(128,))
 
-assert (cl.from_device(d_out) == input + 5).all()
+assert (cl.from_device(d_out) == input * 2).all()

### 2013-11-14 ace-pldi14/ace-pldi14.tex
Fixed figures 1-5

@@
 Decorators in Python are simply syntactic sugar for applying the decorator function directly to the function  being decorated \cite{python}.
-In other words, line 3 could be replaced by inserting the following  statement on line 9:
+In other words, line 3 could be replaced by inserting the following  statement on line 9\todo{figure out how to center this}:
 
@@
 Generic functions can be generated directly from ASTs as well, providing Ace with support for  straightforward metaprogramming.
-Listing \ref{metaprogramming} shows how to generate two more generic functions, \verb|thresh_scale| and \verb|thresh_dbl|.
+Listing \ref{metaprogramming} shows how to generate two more generic functions, \verb|scale| and \verb|double|.
 The latter is derived from the former using a library for manipulating Python syntax trees, \verb|astx|.
-In particular, its \verb|specialize| function replaces uses of the second argument of \verb|thresh_scale| with the literal \verb|2.0| (and changes the function's name), leaving a function of one argument.
+In particular, its \verb|specialize| function replaces uses of the second argument of \verb|scale| with the literal \verb|2| (and changes the function's name), leaving a function of one argument.
 
@@
 Hello, compile-time world!
+[ace] TypeError in listing1.py (line 6, col 28): 
+      'GenericFnType(negate)' does not support [].
 [acec] listing3.cl successfully generated.

### 2013-11-14 ace-pldi14/listing4.py

@@ -1,8 +1,10 @@
-import listing1, listing2, examples.clx as clx
+import listing1, listing2, ace, examples.clx as clx
 
 T1 = clx.Ptr(clx.global_, clx.float)
 T2 = clx.Ptr(clx.global_, clx.cplx(clx.int))
-TF = listing2.thresh_dbl.ace_type
+TF = listing2.negate.ace_type
 
-map_add5_float = listing1.map[[T1, T1, TF]]
-map_add5_c_int = listing1.map[[T2, T2, TF]]
+try: map_neg_f32 = listing1.map[[TF, T1, T1]]
+except ace.TypeError as e: print e.full_msg
+map_neg_f32 = listing1.map[[T1, T1, TF]]
+map_neg_ci32 = listing1.map[[T2, T2, TF]]

### 2013-11-14 ace-pldi14/listing7.py

@@ -1,9 +1,9 @@
 import ace, examples.clx as clx, ast, astx
 
-cl_fn = ace.fn(clx.base, clx.opencl)
-thresh_scale = cl_fn(ast.parse("""
-def thresh_scale(x, s):
-	return x * s if x >= 0 else 0"""))
+_fn = ace.fn(clx.base, clx.opencl)
 
-thresh_dbl = cl_fn(astx.specialize(thresh_scale.ast, 
-	"thresh_dbl", s=ast.parse("2.0"))
+scale = _fn(ast.parse("""def scale(x, s): 
+	return x * s""")
+
+negate = _fn(astx.specialize(scale.ast, "negate", 
+	s=ast.parse("-1"))

### 2013-11-14 ace-pldi14/ace-pldi14.tex
More progress on Sec. 2

@@
 Generic functions can be generated directly from ASTs as well, providing Ace with support for  straightforward metaprogramming.
-Listing \ref{metaprogramming} shows how to generate two more generic functions, \verb|scale| and \verb|double|.
-The latter is derived from the former using a library for manipulating Python syntax trees, \verb|astx|.
-In particular, its \verb|specialize| function replaces uses of the second argument of \verb|scale| with the literal \verb|2| (and changes the function's name), leaving a function of one argument.
+Listing \ref{metaprogramming} shows how to generate two more generic functions, \verb|scale| and \verb|negate|.
+The latter is derived from the former by using a library for manipulating Python syntax trees, \verb|astx|.
+In particular, the \verb|specialize| function replaces uses of the second argument of \verb|scale| with the literal \verb|-1| (and changes the function's name), leaving a function of one argument.
+
+\subsection{Explicit Type Assignment}
+To compile a generic function to a particular \emph{concrete function}, a type must be provided for each argument, so that typechecking and translation can proceed.
+Listing \ref{compscript} shows how to explicitly provide type assignments to \verb|map| via the subscript operator.
+We attempt to do so three times in Listing \ref{compscript}.
+The first, on line \ref{compscript}.7, fails due to a type error, which we handle so that the script can proceed.
+The error occurred  because the ordering of types was incorrect.
+We provide a valid ordering on line \ref{compscript}.9 to generate the concrete function \verb|map_neg_f32|.
+We then provide a different type assignment to generate the concrete function \verb|map_neg_ci32|.
+Concrete functions are instances of \verb|ace.ConcreteFn|, consisting of an abstract syntax tree annotated with types and translations along with a reference to the original generic function.
+%The typing and translation process is mediated by the logic in the active base, types and target that have been provided, as we will describe in more detail below.
+
+To produce an output file from an Ace compilation script like \verb|listing|\texttt{\ref{compscript}}\verb|.py|, the command \verb|acec| can be invoked from the shell, as shown in Listing \ref{mapc}.
+The \verb|acec| compiler operates in two steps:
+\begin{enumerate}
+\item Executes the provided Python file (\verb|listing3.py|).
+\item Extracts the translation from concrete functions and other top-level constructs (e.g.
+types requiring declarations, or generated imports and pragmas) in the top-level Python environment.
+This may produce one or more files (here, just \verb|listing3.cl|) as determined by the active targets that were used (for example, a web framework built upon Ace might produce separate HTML, CSS and JavaScript files; see Sec.
+\ref{active-targets}).
+\end{enumerate}
+
+In this case, step 1 prints the output on lines \ref{mapc}.2-\ref{mapc}.4.
+The type error printed on lines \ref{mapc}.3-\ref{mapc}.4 will be explained in the next section.
+The compiler (which is a simple Python script) concludes with the message on line \ref{mapc}.5 to indicate that one file was generated.
+This file is shown in Listing \ref{mapout} and can be used by any programs that consume OpenCL code (e.g.
+a C file that invokes the generated kernels via the OpenCL host API).
+We will show in Section \ref{backend} that for targets with Python bindings, such as OpenCL, CUDA, C, Java or Python itself, generic functions can be executed directly, without any of the explicit compilation steps in Listings \ref{compscript}-\ref{mapout}.
 
 \subsection{Types}
-To go from a generic function to a \emph{concrete function}, we must provide a type assignment for its arguments.
-We will show different ways to do this in Secs.
-\ref{procedural} and \ref{implicit}.
-Before we can do this, however, we must understand what types are.
+\begin{codelisting}
+\lstinputlisting{listing4.py}
+\caption{[\texttt{listing\ref{compscript}.py}] The generic \texttt{map} function compiled to map the \texttt{negate} function over two  types of input.}
+\label{compscript}
+\end{codelisting}
+\begin{codelisting}
+\begin{lstlisting}[style=Bash]
+$ `acec listing3.py`
+Hello, compile-time world!
+[ace] TypeError in listing1.py (line 6, col 28): 
+      'GenericFnType(negate)' does not support [].
+[acec] listing3.cl successfully generated.
+\end{lstlisting}
+\caption{Compiling \texttt{listing\ref{compscript}.py} using the \texttt{acec} compiler.}
+\label{mapc}
+\end{codelisting}
+\begin{codelisting}
+\lstinputlisting[style=OpenCL]{listing5.cl}
+\caption{[\texttt{listing\ref{compscript}.cl}] The OpenCL file generated by Listing \ref{mapc}.}
+\label{mapout}
+\end{codelisting}
+Lines 3-5 of Listing \ref{compscript} construct the types used on lines 7-10.
+In Ace, types are themselves values that can be manipulated at compile-time, in contrast to most  contemporary languages, where user-defined types are simply declared at compile-time but cannot be defined, inspected or passed around programmatically.
+More specifically, types are instances of a Python class that implements the \verb|ace.ActiveType| interface (see Sec.
+\ref{extensions}).
+Python, then, serves as the \emph{type-level language} of Ace.
+Being Python values, types can be assigned to variables for convenience (a construct similar to \verb|typedef| in C or \verb|type| in Haskell).
+Types, like all compile-time mechanisms in Ace, do not have mutable state and operate in a referentially transparent manner.% These types are all implemented in the \verb|clx| library imported on line 1, none are built into Ace itself.
 
-In Ace, types are themselves values that can be manipulated at compile-time (unlike most contemporary languages, where user-defined types are simply declared at compile-time and cannot be defined or manipulated programmatically).
-In particular, types are instances of a Python class that implements the \verb|ace.ActiveType| interface.
-Python, then, serves also as the \emph{type-level language} of Ace (see related work in Sec.
-\ref{related}).
-
-Listing 2 shows three types, on lines 7-9, being assigned to variables for convenience (a construct similar to \verb|typedef| in C-like languages or \verb|type| in Haskell).
-These types all come from the \verb|clx| library imported on line 1, none are built into Ace itself.
-
-The type named \verb|F| on line 7 corresponds to a type in OpenCL, \verb|global float*|: a pointer to a 32-bit floating point number stored in the compute device's global memory (one of four memory spaces defined by OpenCL, see \cite{opencl11}).
-Here, \verb|clx.float| is an Ace type corresponding to \verb|float| in OpenCL, and \verb|clx.Ptr| is an Ace type family corresponding to pointer types, indexed by both an address space (here, \verb|clx.global_|) and the type being pointed to.
-The \verb|clx| library contains a complete implementation of the OpenCL type system, which is based on C99.
-We will return to how this works in Section \ref{att}.
-We note that we are being more verbose than necessary for pedagogical reasons.
-The \verb|clx| library defines more concise naming schemas for OpenCL's built-in types: \verb|F| could be written \verb|clx.gp(clx.f32)|.
+The type named \verb|T1| on line 7 corresponds to the OpenCL type \verb|global float*|: a pointer to a 32-bit floating point number stored in the compute device's global memory (one of four memory spaces defined by OpenCL \cite{opencl11}).
+It is constructed by applying \verb|clx.Ptr|, which is an Ace type constructor corresponding to pointer types, to an address space (here, \verb|clx.global_|) and the type being pointed to, \verb|clx.float|, the Ace type corresponding to \verb|float| in OpenCL.
+The \verb|clx| library contains a full implementation of the OpenCL type system.
+We will return to how types are implemented in Section \ref{att}.
+We note that we are being more verbose than necessary for the sake of pedagogy.
+The \verb|clx| library includes a concise shorthand for OpenCL's built-in types: \verb|T1| could be written \verb|clx.gp(clx.f32)|.
 We also note that Ace is \emph{unopinionated} about issues like memory safety.
 We will show how to implement safer alternatives to raw pointers in Sec.
-\ref{examples}, but we allow users to choose a low level of abstraction when they wish.(see Sec.
-\ref{discussion}).
+\ref{examples}, but we allow users to choose a low level of abstraction if they wish (e.g.
+for compatibility with existing code).
 
-Listing \ref{mapadd5dbl} shows how to assign types to the arguments of \verb|map| so that this logic can be typechecked and compiled.
+The type \verb|T2| on line 8 does not correspond to an OpenCL type directly.
+Rather, it is a global pointer to a user-defined primitive type representing complex numbers (integers, in this case).
+OpenCL does not include primitive support for complex numbers (as of this writing)
 
@@
 To create a {\em concrete function} (that is, an instance of the class \verb|ace.ConcreteFn|) that can be emitted by the compiler, types must be assigned to each of the arguments.
-Listing \ref{mapadd5dbl} shows how to use the \verb|compile| method to specialize \verb|map| in \emph{two} different ways to apply the \verb|add5| function, defined on Lines 4-6, to arrays  that reside in global memory (OpenCL associates a memory space with pointer types).
+Listing \ref{compscript} shows how to use the \verb|compile| method to specialize \verb|map| in \emph{two} different ways to apply the \verb|add5| function, defined on Lines 4-6, to arrays  that reside in global memory (OpenCL associates a memory space with pointer types).
 Line 9 results in a version specialized for arrays of \verb|double|s and Line 10 results in a version for arrays of \verb|int|s.
@@
 \subsection{Programmatic Compilation}\label{procedural}
-\begin{codelisting}
-\lstinputlisting[linebackgroundcolor={\btLstHL{5-5}}]{listing4.py}
-\caption{[\texttt{listing\ref{mapadd5dbl}.py}] The generic \texttt{map} function compiled to map the \texttt{add\_5} function over two  types of input.}
-\label{mapadd5dbl}
-\end{codelisting}
-
-\begin{codelisting}
-\begin{lstlisting}[style=Bash]
-$ `acec listing3.py`
-Hello, compile-time world!
-[ace] TypeError in listing1.py (line 6, col 28): 
-      'GenericFnType(negate)' does not support [].
-[acec] listing3.cl successfully generated.
-\end{lstlisting}
-\caption{Compiling \texttt{listing\ref{mapadd5dbl}.py} using the \texttt{acec} compiler.}
-\label{mapc}
-\end{codelisting}
-\begin{codelisting}
-\lstinputlisting[style=OpenCL]{listing5.cl}
-\caption{[\texttt{listing\ref{mapadd5dbl}.cl}] The OpenCL file generated by Listing \ref{mapc}.}
-\label{mapout}
-\end{codelisting}
 As defined on Lines 5-7, \verb|main| is a \emph{generic function}.
@@
 
-Before moving on to more interesting examples, let us discuss the \verb|acec| compiler shown operating at the shell in Listing \ref{helloout}. The \verb|acec| compiler operates in two steps:
-\begin{enumerate}
-\item Executes the provided Python file (\verb|hello.py|) which contains Ace functions and (in future examples) types and compile-time code generation logic.
-\item Produces source code for concrete functions (produced using the \verb|compile| method) in the top-level Python environment and any other concrete functions, type declarations and other program items required by or generated by these functions.
-This may produce one or more files (here, just \verb|hello.cl|).
-\end{enumerate}
-
-We will show in Section \ref{backend} that for backends with Python bindings, such as OpenCL, CUDA and C, generic functions can be executed directly, without this explicit compilation step to concrete functions, if desired.
+Before moving on to more interesting examples, let us discuss the \verb|acec| compiler shown operating at the shell in Listing \ref{helloout}.
 
@@
 Let us again consider the data-parallel map example of Section 2.
-When \verb|map| is compiled on Line 9 of Listing \ref{mapadd5dbl}, the first two argument types are \verb|gptr(double)|.
+When \verb|map| is compiled on Line 9 of Listing \ref{compscript}, the first two argument types are \verb|gptr(double)|.
 This type is an instance of the user-defined \verb|OpenCL.PtrType| which inherits from \verb|ace.Type|.
@@
 
+ Future work: integrate with something like scalad http://lampwww.epfl.ch/~hmiller/scala2013/resources/pdfs/paper8.pdf.
+
 %\acks

### 2013-11-14 ace-pldi14/ace-pldi14.tex
pldi14 2.5 p1

@@
 
-\subsection{Explicit Type Assignment}
+\subsection{Concrete Functions and Explicit Compilation}
 To compile a generic function to a particular \emph{concrete function}, a type must be provided for each argument, so that typechecking and translation can proceed.
@@
 a C file that invokes the generated kernels via the OpenCL host API).
-We will show in Section \ref{backend} that for targets with Python bindings, such as OpenCL, CUDA, C, Java or Python itself, generic functions can be executed directly, without any of the explicit compilation steps in Listings \ref{compscript}-\ref{mapout}.
+We will show in Section \ref{backend} that for targets with Python bindings, such as OpenCL, CUDA, C, Java or Python itself, generic functions can be executed directly, without any of the explicit compilation steps in Listings \ref{compscript}-\ref{mapc}.
 
@@
 \end{codelisting}
-Lines 3-5 of Listing \ref{compscript} construct the types used on lines 7-10.
-In Ace, types are themselves values that can be manipulated at compile-time, in contrast to most  contemporary languages, where user-defined types are simply declared at compile-time but cannot be defined, inspected or passed around programmatically.
+Lines \ref{compscript}.3-\ref{compscript}.5 construct the types assigned on lines \ref{compscript}.7-\ref{compscript}.10.
+In Ace, types are themselves values that can be manipulated at compile-time, in contrast to most  contemporary languages, where user-defined types are declaratively written at compile-time but cannot be constructed, inspected or passed around programmatically.
 More specifically, types are instances of a Python class that implements the \verb|ace.ActiveType| interface (see Sec.
@@
 Being Python values, types can be assigned to variables for convenience (a construct similar to \verb|typedef| in C or \verb|type| in Haskell).
-Types, like all compile-time mechanisms in Ace, do not have mutable state and operate in a referentially transparent manner.% These types are all implemented in the \verb|clx| library imported on line 1, none are built into Ace itself.
+Types, like all compile-time mechanisms in Ace (functions, bases, etc.), do not have mutable state and operate in a referentially transparent manner.% These types are all implemented in the \verb|clx| library imported on line 1, none are built into Ace itself.
 
-The type named \verb|T1| on line 7 corresponds to the OpenCL type \verb|global float*|: a pointer to a 32-bit floating point number stored in the compute device's global memory (one of four memory spaces defined by OpenCL \cite{opencl11}).
+The type named \verb|T1| on line \ref{compscript}.7 corresponds to the OpenCL type \verb|global float*|: a pointer to a 32-bit floating point number stored in the compute device's global memory (one of four memory spaces defined by OpenCL \cite{opencl11}).
 It is constructed by applying \verb|clx.Ptr|, which is an Ace type constructor corresponding to pointer types, to an address space (here, \verb|clx.global_|) and the type being pointed to, \verb|clx.float|, the Ace type corresponding to \verb|float| in OpenCL.
@@
 This example demonstrates several key design decisions that characterize Ace: static typing of run-time behavior, a Python-based syntax, a phase distinction between compile-time and run-time logic, and programmatic
-\subsection{Programmatic Compilation}\label{procedural}
-As defined on Lines 5-7, \verb|main| is a \emph{generic function}.
-This terminology in Ace is used to indicate a function for which types have not yet been assigned to arguments.
-Invoking the \verb|compile| method on a generic function with a sequence of argument types invokes the active typechecking and translation mechanism we will describe in Section 3 to produce a \emph{concrete function} -- one with a single type for each argument, internal variable (such as \verb|hello|) and the return value.
-In Listing \ref{hello}, we name this concrete function \verb|main|, overwriting the generic function of the same name that it is derived from.
-The compiler does nothing apparently interesting: there are no arguments, the single internal variable takes the OpenCL string type \verb|char*| from its value and the return type is \verb|void|.
 
-Before moving on to more interesting examples, let us discuss the \verb|acec| compiler shown operating at the shell in Listing \ref{helloout}.
-
-These types are \emph{objects in the metalanguage}, Python.
-More specifically, types are {instances} of user-defined classes that inherit from the Ace-provided \verb|ace.Type| class.
-For example, \verb|gptr(double)| is an instance of \verb|OpenCL.PtrType| instantiated with the target type, \verb|double|, and memory space, \verb|__global|, as constructor arguments.
-The types \verb|double| and \verb|int| are instances of \verb|OpenCL.FloatType| and \verb|OpenCL.IntegerType|, respectively.
 \subsection{Type Propagation}

### 2013-11-14 ace-pldi14/ace-pldi14.tex
pldi14 2.5 p2+3

@@
 This may produce one or more files (here, just \verb|listing3.cl|) as determined by the active targets that were used (for example, a web framework built upon Ace might produce separate HTML, CSS and JavaScript files; see Sec.
-\ref{active-targets}).
+\ref{atargets}).
 \end{enumerate}
@@
 More specifically, types are instances of a Python class that implements the \verb|ace.ActiveType| interface (see Sec.
-\ref{extensions}).
+\ref{atypes}).
 Python, then, serves as the \emph{type-level language} of Ace.
-Being Python values, types can be assigned to variables for convenience (a construct similar to \verb|typedef| in C or \verb|type| in Haskell).
-Types, like all compile-time mechanisms in Ace (functions, bases, etc.), do not have mutable state and operate in a referentially transparent manner.% These types are all implemented in the \verb|clx| library imported on line 1, none are built into Ace itself.
+As Python values, types can be assigned to variables when desired (removing the need for  facilities like \verb|typedef| in C or \verb|type| in Haskell).
+Types, like all compile-time objects derived from Ace base classes, do not have visible state and operate in a referentially transparent manner (by constructor memoization, which we do not detail here).% These types are all implemented in the \verb|clx| library imported on line 1, none are built into Ace itself.
 
-The type named \verb|T1| on line \ref{compscript}.7 corresponds to the OpenCL type \verb|global float*|: a pointer to a 32-bit floating point number stored in the compute device's global memory (one of four memory spaces defined by OpenCL \cite{opencl11}).
-It is constructed by applying \verb|clx.Ptr|, which is an Ace type constructor corresponding to pointer types, to an address space (here, \verb|clx.global_|) and the type being pointed to, \verb|clx.float|, the Ace type corresponding to \verb|float| in OpenCL.
+The type named \verb|T1| on line \ref{compscript}.3 corresponds to the OpenCL type \verb|global float*|: a pointer to a 32-bit floating point number stored in the compute device's global memory (one of four memory spaces defined by OpenCL \cite{opencl11}).
+It is constructed by applying \verb|clx.Ptr|, which is an Ace type constructor corresponding to pointer types, to a value representing an  address space, \verb|clx.global_|, and the type being pointed to.
+That type, \verb|clx.float|, is in turn the Ace type corresponding to \verb|float| in OpenCL (which, unlike C99, is always 32 bits).
 The \verb|clx| library contains a full implementation of the OpenCL type system.
-We will return to how types are implemented in Section \ref{att}.
 We note that we are being more verbose than necessary for the sake of pedagogy.
@@
 We will show how to implement safer alternatives to raw pointers in Sec.
-\ref{examples}, but we allow users to choose a low level of abstraction if they wish (e.g.
-for compatibility with existing code).
+\ref{examples}\todo{true?}, but we allow users to choose a low level of abstraction if the need arises.
 
-The type \verb|T2| on line 8 does not correspond to an OpenCL type directly.
-Rather, it is a global pointer to a user-defined primitive type representing complex numbers (integers, in this case).
-OpenCL does not include primitive support for complex numbers (as of this writing)
+The type \verb|T2| on line \ref{compscript}.4 does not correspond to type in OpenCL directly.
+It is a pointer to a \emph{complex integer} in global memory.
+OpenCL does not include primitive support for complex numbers, however -- it is an extension available only in \verb|clx|.
+The \verb|clx.Cplx| type constructor is parameterized by the numeric type that should be used for the real and imaginary parts, here \verb|clx.int|, the type corresponding to 32-bit OpenCL integers.
+Arithmetic is supported with other types in the same family, as well as with plain numeric types (treated as if their imaginary part was zero).
+When targeting OpenCL, Ace expressions of type \verb|clx.Cplx(clx.int)| are compiled to OpenCL expressions of type \verb|int2|, a pair of 32-bit integers.
+This can be observed in several places on lines \ref{mapout}.14-\ref{mapout}.21.
+This is merely an implementation detail of \verb|clx.Cplx|, however.
+An Ace value of type \verb|clx.int2| can\emph{not} be used when a \verb|clx.Cplx(clx.int)| is expected (attempting to do so will result in a static type error).
+That is, \verb|clx.Cplx| truly extends the type system, it is not merely a type alias (see Sec.
+\ref{atypes}).
+
+The type \verb|TF| on line \ref{compscript}.5
 
 This notion of types as metalanguage objects is key to the Ace compilation model and also enables other mechanisms that we will discuss in subsequent sections.
-
-Running \verb|acec listing3.py| would produce no output.
-To create a {\em concrete function} (that is, an instance of the class \verb|ace.ConcreteFn|) that can be emitted by the compiler, types must be assigned to each of the arguments.
-Listing \ref{compscript} shows how to use the \verb|compile| method to specialize \verb|map| in \emph{two} different ways to apply the \verb|add5| function, defined on Lines 4-6, to arrays  that reside in global memory (OpenCL associates a memory space with pointer types).
-Line 9 results in a version specialized for arrays of \verb|double|s and Line 10 results in a version for arrays of \verb|int|s.
-The output of compilation is shown in Listing \ref{mapout}.
-
-The \verb|map| function defined in Listing \ref{map} will compile to one or more kernels that applies a transfer function, \verb|f|, to the element of the input array, \verb|input|, corresponding to its global id.
-It writes the result of this call into the corresponding location in the provided output array, \verb|output|.
-If the kernel is running in thread 0, it prints a message.
-The functions \verb|get_global_id| and \verb|printf| are OpenCL primitive functions.
-To emphasize again that the semantics of this code differs from Python's, we note that these functions have no meaning in Python.
-
-%\begin{codelisting}
-%\lstinputlisting{hello.py}
-%\caption{\texttt{[hello.py]} A basic Ace program demonstrating the two-phase structure of Ace programs and libraries.}
-%\label{hello}
-%\end{codelisting}
-%A variant of the standard ``Hello, World!'' example written in Ace is shown in Listing \ref{hello} and its compilation to statically-typed OpenCL kernel code is demonstrated in Listing \ref{helloout}.
-The OpenCL kernel language is a variant of C99 with some additions and restrictions to facilitate execution on GPUs and other accelerators, in addition to conventional CPUs.
-We will discuss it further in subsequent examples.
-%We emphasize that this module, which we use throughout the paper, is simply a library like any other.
-The core of Ace gives no special treatment to it; it is distributed together with Ace for convenience.
-Aspects of its implementation will be described  in Section \ref{att}.
-
-This example demonstrates several key design decisions that characterize Ace: static typing of run-time behavior, a Python-based syntax, a phase distinction between compile-time and run-time logic, and programmatic
 
@@
 \section{Extensibility}\label{att}
+\subsection{Active Bases}\label{abases}
+\subsection{Active Types}\label{atypes}
+\subsection{Active Targets}\label{atargets}
+
 If no active target is explicitly provided, as in our example, the active base  provides a default target.

### 2013-11-14 ace-pldi14/listing4.py

@@ -1,7 +1,7 @@
 import listing1, listing2, ace, examples.clx as clx
 
 T1 = clx.Ptr(clx.global_, clx.float)
-T2 = clx.Ptr(clx.global_, clx.cplx(clx.int))
+T2 = clx.Ptr(clx.global_, clx.Cplx(clx.int))
 TF = listing2.negate.ace_type
 
 try: map_neg_f32 = listing1.map[[TF, T1, T1]]

### 2013-11-14 ace-pldi14/ace-pldi14.tex
pldi14 2.5 p2+3 updated

@@
 
-\subsection{Generic Functions}
+\subsection{Generic Functions}\label{genfn}
 \begin{codelisting}
@@
 \end{codelisting}
-Lines 3-4 introduce \verb|map|, an Ace function governed by the active base \verb|clx.base| taking three arguments.
-The \emph{active target}, \verb|clx.opencl|, determines which language the function will target (here, the OpenCL kernel language) and mediates code generation.
+Lines 3-4 introduce \verb|map|, an Ace function of three arguments that is governed by the \emph{active base} referred to by \verb|clx.base| and targeting the \emph{active target} referred to by \verb|clx.opencl|.
+The active target determines which language the function will compile to (here, the OpenCL kernel language) and mediates code generation.
 
@@
 Instead, it will be governed by the active base together with the active types used within it.
-No such types have been provided explicitly for the arguments yet, however.
-Because our type system is extensible, the code inside could be meaningful for many different assignments of types to the arguments, so \verb|map| is not yet a \emph{concrete function}.
+No such types have been provided explicitly, however.
+Because our type system is extensible, the code inside could be meaningful for many different assignments of types to the arguments.
 We call functions awaiting types \emph{generic functions}.
-%Once types have been assigned, they are called \emph{concrete functions}.
+Once types have been assigned, they are called \emph{concrete functions}.
 
@@
 \subsection{Concrete Functions and Explicit Compilation}
-To compile a generic function to a particular \emph{concrete function}, a type must be provided for each argument, so that typechecking and translation can proceed.
-Listing \ref{compscript} shows how to explicitly provide type assignments to \verb|map| via the subscript operator.
+To compile a generic function to a particular \emph{concrete function}, a type must be provided for each argument, and typechecking and translation must then succeed.
+Listing \ref{compscript} shows how to explicitly provide type assignments to \verb|map| using the subscript operator.
 We attempt to do so three times in Listing \ref{compscript}.
 The first, on line \ref{compscript}.7, fails due to a type error, which we handle so that the script can proceed.
-The error occurred  because the ordering of types was incorrect.
+The error occurred  because the ordering of the argument types was incorrect.
 We provide a valid ordering on line \ref{compscript}.9 to generate the concrete function \verb|map_neg_f32|.
@@
 
-To produce an output file from an Ace compilation script like \verb|listing|\texttt{\ref{compscript}}\verb|.py|, the command \verb|acec| can be invoked from the shell, as shown in Listing \ref{mapc}.
-The \verb|acec| compiler operates in two steps:
+To produce an output file from an Ace ``compilation script'' like \verb|listing|\texttt{\ref{compscript}}\verb|.py|, the command \verb|acec| can be invoked from the shell, as shown in Listing \ref{mapc}.
+The \verb|acec| compiler (a simple Python script) operates in two stages:
 \begin{enumerate}
 \item Executes the provided Python file (\verb|listing3.py|).
-\item Extracts the translation from concrete functions and other top-level constructs (e.g.
+\item Extracts the translations from concrete functions and other top-level constructs (e.g.
 types requiring declarations, or generated imports and pragmas) in the top-level Python environment.
-This may produce one or more files (here, just \verb|listing3.cl|) as determined by the active targets that were used (for example, a web framework built upon Ace might produce separate HTML, CSS and JavaScript files; see Sec.
+This may produce one or more files as mediated by the active targets that were used (here, just \verb|listing3.cl|, but a web framework built upon Ace might produce separate HTML, CSS and JavaScript files; see Sec.
 \ref{atargets}).
@@
 
-In this case, step 1 prints the output on lines \ref{mapc}.2-\ref{mapc}.4.
+In this case, stage 1 results in the output on lines \ref{mapc}.2-\ref{mapc}.4.
 The type error printed on lines \ref{mapc}.3-\ref{mapc}.4 will be explained in the next section.
-The compiler (which is a simple Python script) concludes with the message on line \ref{mapc}.5 to indicate that one file was generated.
+The compiler then enters stage 2 and concludes with the message on line \ref{mapc}.5 to indicate that one file was generated.
 This file is shown in Listing \ref{mapout} and can be used by any programs that consume OpenCL code (e.g.
-a C file that invokes the generated kernels via the OpenCL host API).
+a C program that invokes the generated kernels via the OpenCL host API).
 We will show in Section \ref{backend} that for targets with Python bindings, such as OpenCL, CUDA, C, Java or Python itself, generic functions can be executed directly, without any of the explicit compilation steps in Listings \ref{compscript}-\ref{mapc}.
@@
 \end{codelisting}
-Lines \ref{compscript}.3-\ref{compscript}.5 construct the types assigned on lines \ref{compscript}.7-\ref{compscript}.10.
-In Ace, types are themselves values that can be manipulated at compile-time, in contrast to most  contemporary languages, where user-defined types are declaratively written at compile-time but cannot be constructed, inspected or passed around programmatically.
+Lines \ref{compscript}.3-\ref{compscript}.5 construct the types assigned to the arguments of \verb|map| on lines \ref{compscript}.7-\ref{compscript}.10.
+In Ace, types are themselves values that can be manipulated at compile-time.
+This stands in contrast to other contemporary languages, where user-defined types (e.g.
+datatypes, classes, structs) are written declaratively at compile-time but cannot be constructed, inspected or passed around programmatically.
 More specifically, types are instances of a Python class that implements the \verb|ace.ActiveType| interface (see Sec.
 \ref{atypes}).
-Python, then, serves as the \emph{type-level language} of Ace.
-As Python values, types can be assigned to variables when desired (removing the need for  facilities like \verb|typedef| in C or \verb|type| in Haskell).
+As Python values, types can be assigned to variables when convenient (removing the need for  facilities like \verb|typedef| in C or \verb|type| in Haskell).
 Types, like all compile-time objects derived from Ace base classes, do not have visible state and operate in a referentially transparent manner (by constructor memoization, which we do not detail here).% These types are all implemented in the \verb|clx| library imported on line 1, none are built into Ace itself.
 
-The type named \verb|T1| on line \ref{compscript}.3 corresponds to the OpenCL type \verb|global float*|: a pointer to a 32-bit floating point number stored in the compute device's global memory (one of four memory spaces defined by OpenCL \cite{opencl11}).
-It is constructed by applying \verb|clx.Ptr|, which is an Ace type constructor corresponding to pointer types, to a value representing an  address space, \verb|clx.global_|, and the type being pointed to.
+The type named \verb|T1| on line \ref{compscript}.3 corresponds to the OpenCL type \verb|global float*|: a pointer to a 32-bit floating point number stored in the compute device's global memory (one of four address spaces defined by OpenCL \cite{opencl11}).
+It is constructed by applying \verb|clx.Ptr|, which is an Ace type constructor corresponding to pointer types, to a value representing the  address space, \verb|clx.global_|, and the type being pointed to.
 That type, \verb|clx.float|, is in turn the Ace type corresponding to \verb|float| in OpenCL (which, unlike C99, is always 32 bits).
-The \verb|clx| library contains a full implementation of the OpenCL type system.
-We note that we are being more verbose than necessary for the sake of pedagogy.
-The \verb|clx| library includes a concise shorthand for OpenCL's built-in types: \verb|T1| could be written \verb|clx.gp(clx.f32)|.
-We also note that Ace is \emph{unopinionated} about issues like memory safety.
-We will show how to implement safer alternatives to raw pointers in Sec.
-\ref{examples}\todo{true?}, but we allow users to choose a low level of abstraction if the need arises.
+The \verb|clx| library contains a full implementation of the OpenCL type system (including behaviors, like promotions, inherited from C99).
+Ace is \emph{unopinionated} about issues like memory safety and the wisdom of such promotions.
+We will discuss how to implement, as libraries, abstractions that are higher-level than raw pointers in Sec.
+\ref{examples}, but Ace does not prevent users from choosing a low level of abstraction or ``interesting'' semantics if the need arises (e.g.
+for compatibility with existing libraries; see the discussion in Sec.
+\ref{discussion}).
+We also note that we are being more verbose than necessary for the sake of pedagogy.
+The \verb|clx| library includes more concise shorthand for OpenCL's types: \verb|T1| is equal to \verb|clx.gp(clx.f32)|.
+%Similarly, the decorators in Listings \ref{map} and \ref{metaprogramming} could have been written \verb|clx.cl_fn|.\todo{move this back there probably}
 
-The type \verb|T2| on line \ref{compscript}.4 does not correspond to type in OpenCL directly.
-It is a pointer to a \emph{complex integer} in global memory.
-OpenCL does not include primitive support for complex numbers, however -- it is an extension available only in \verb|clx|.
-The \verb|clx.Cplx| type constructor is parameterized by the numeric type that should be used for the real and imaginary parts, here \verb|clx.int|, the type corresponding to 32-bit OpenCL integers.
-Arithmetic is supported with other types in the same family, as well as with plain numeric types (treated as if their imaginary part was zero).
-When targeting OpenCL, Ace expressions of type \verb|clx.Cplx(clx.int)| are compiled to OpenCL expressions of type \verb|int2|, a pair of 32-bit integers.
+The type \verb|T2| on line \ref{compscript}.4 is a pointer to a \emph{complex integer} in global memory.
+It does not correspond direrctly to a type in OpenCL, because OpenCL does not include primitive support for complex numbers.
+Instead, it uses an active type constructor \verb|clx.Cplx|, which includes the necessary logic for typechecking operations on complex numbers and translating them to OpenCL (Sec.
+\ref{atypes}).
+This constructor is parameterized by the numeric type that should be used for the real and imaginary parts, here \verb|clx.int|, which corresponds to 32-bit OpenCL integers.
+Arithmetic operations with other complex numbers, as well as with plain numeric types (treated as if their imaginary part was zero), are supported.
+When targeting OpenCL, Ace expressions assigned type \verb|clx.Cplx(clx.int)| are compiled to OpenCL expressions of type \verb|int2|, a  \emph{vector type} of two 32-bit integers (a type that itself is not inherited from C99).
 This can be observed in several places on lines \ref{mapout}.14-\ref{mapout}.21.
-This is merely an implementation detail of \verb|clx.Cplx|, however.
-An Ace value of type \verb|clx.int2| can\emph{not} be used when a \verb|clx.Cplx(clx.int)| is expected (attempting to do so will result in a static type error).
-That is, \verb|clx.Cplx| truly extends the type system, it is not merely a type alias (see Sec.
-\ref{atypes}).
+This choice is merely an implementation detail that can be kept private to \verb|clx|, however.
+An Ace value of type \verb|clx.int2| (that is, an actual OpenCL vector) \emph{cannot} be used when a \verb|clx.Cplx(clx.int)| is expected (and attempting to do so will result in a static type error).% \verb|clx.Cplx| truly extends the type system, it is not a type alias.
 
-The type \verb|TF| on line \ref{compscript}.5
+The type \verb|TF| on line \ref{compscript}.5 is extracted from the generic function \verb|negate| constructed in Listing \ref{metaprogramming}.
+Generic functions, according to Sec.
+\ref{genfn}, have not yet had a type assigned to them, so it may seem perplexing that we are nevertheless assigning a type to it.
 

### 2013-11-14 ace-pldi14/ace-pldi14.tex
Got rid of a stupid todo, centered verbatim

--- 2013-11-14 ace-pldi14/ace-pldi14.tex
@@
 \usepackage{ stmaryrd }
+\usepackage{verbatimbox}
 
@@
 Decorators in Python are simply syntactic sugar for applying the decorator function directly to the function  being decorated \cite{python}.
-In other words, line 3 could be replaced by inserting the following  statement on line 9\todo{figure out how to center this}:
+In other words, line 3 could be replaced by inserting the following  statement on line 9:
 
-\verb|map = ace.fn(clx.base, clx.opencl)(map)|
+\vspace{-0.2cm}
+\begin{verbbox}
+map = ace.fn(clx.base, clx.opencl)(map)
+\end{verbbox}
+\begin{figure}[h!]
+  \centering
+  \theverbbox
+\end{figure}
+\vspace{-0.2cm}
 

### 2013-11-14 ace-pldi14/ace-pldi14.tex
Makes columns in table two lines

--- 2013-11-14 ace-pldi14/ace-pldi14.tex
@@
 Active Types & Ace & x & ~ & x & x & x\\
-Desugaring & SugarJ, Sugar* & ~ & x & ~ & ~ & ~\\
-Rule Injection & Qi, Typed Racket/Clojure, A? & 1+2 & ~ & x & ~ & ~\\
-Static metaprogramming & Scala macros, OJ, OC++ & ~ & ~ & ~ & x & ~\\
-Cross-compilation & Delite & x & ~ & ~ & ~ & x\\
+Desugaring & SugarJ \cite{erdweg2011sugarj}, Sugar* & ~ & x & ~ & ~ & ~\\
+Rule Injection & Qi, Typed Racket \cite{TypedScheme2008}/Clojure, A? & 1+2 & ~ & x & ~ & ~\\
+Static metaprogramming & Scala macros \cite{ScalaMacros2013}, OJ \cite{OpenJava2000}, OC++ \cite{OpenCpp1995}, MorphJ \cite{MorphJ2011} & ~ & ~ & ~ & x & ~\\
+Cross-compilation & Delite \cite{Delite2011} & x & ~ & ~ & ~ & x\\
 EDSL frameworks & ? & x & x & x & ~ & ~\\

### 2013-11-14 ace-pldi14/ace-pldi14.tex
pldi14 2.5 p3+4+5

--- 2013-11-14 ace-pldi14/ace-pldi14.tex
@@
 Generic functions, according to Sec.
-\ref{genfn}, have not yet had a type assigned to them, so it may seem perplexing that we are nevertheless assigning a type to it.
+\ref{genfn}, have not yet had a type assigned to them, so it may seem perplexing that we are nevertheless extracting a type from it.
+Although a conventional arrow type cannot be assigned to \verb|negate|, we can give it a \emph{singleton type}: a type that simply means ``this expression is the \emph{particular} generic function \verb|negate|''.
+This type could also have been explicitly written as \verb|ace.GenericFnType(listing2.negate)|.
+During typechecking and translation of \verb|map_neg_f32| and \verb|map_neg_ci32|, the call to \verb|f| on line \ref{map}.6 operates by using the types of the provided arguments to compile the generic function that inhabits the singleton type of \verb|f| (\verb|negate| in both of these cases) to a concrete function.
+This is why there are two versions of \verb|negate| in the output in Listing \ref{mapout}.
+In other words, types \emph{propagate} into generic functions -- we didn't need to compile \verb|negate| explicitly.
+This also explains the error printed on line \ref{mapc}.3-\ref{mapc}.4: when this type was inadvertently assigned to the first argument \verb|input|, the indexing operation on line \ref{map}.6 resulted in an error.
+A generic function can only be \emph{statically} indexed by a list of types to turn it into a concrete function, not \emph{dynamically} indexed with a value of type \verb|clx.size_t| (the return type of the OpenCL primitive function \verb|get_global_id|).
 
-This notion of types as metalanguage objects is key to the Ace compilation model and also enables other mechanisms that we will discuss in subsequent sections.
+In effect, this scheme enables higher-order functions even when targeting languages, like OpenCL, that have no support for higher-order functions (OpenCL, unlike C99, does not support function pointers).
+Interestingly, because they have a singleton type, they are higher-order but not first-class functions.
+That is, the type system would prevent you from creating a heterogeneous list of generic functions.
+Concrete functions, on the other hand, can be given both a singleton type and a true function type.
+For example, \verb|listing2.scale[[clx.int, clx.int]]| could be given type \verb|ace.Arrow(clx.int, clx.int)|.
+The base determines how to convert the Ace arrow type to an arrow type in the target language (e.g.
+a function pointer for C99, or an integer that indexes into a jump table containing all functions in the program of that type).
 
-\subsection{Type Propagation}
-The type assigned to the third argument, \verb|f|, on both Lines 4.9 and 4.10, is \verb|add5.ace_type|.
-The \verb|ace_type| attribute of a generic function is an instance of \verb|ace.GenericFnType|, the type of Ace generic functions.
-Ace generic functions are compiled to concrete functions automatically at all internal call sites.
-That is, when the compiler encounters the call to \verb|f| inside \verb|map| when compiling \verb|map_add5_double|, it compiles a version of \verb|add5| specialized to the \verb|double| type (seen on Line 5.3), and similarly when compiling \verb|map_add5_int| (on Line 5.16, automatically given a unique name to avoid conflicts).
-This mechanism is called {\em type propagation}.
-We did not need to use \verb|add5.compile(double)| before compiling \verb|map_add5_dbl| because only functions that are never called in the process of compiling other functions in a module need type information explicitly provided, supporting \textbf{ease-of-use} by increasing conciseness.
-
-In effect, this scheme allows for a form of higher-order functional programming even when targeting languages, like OpenCL, that have no support for higher-order functions (OpenCL, unlike C99, does not support function pointers).
-This works because the \verb|ace.GenericFnType| for one function, such as \verb|add5|, is not equal to  the \verb|ace.GenericFnType| for a superficially similar function, such as \verb|add6| (defined as one would expect).
-To put it in type theoretic terms, \verb|ace.GenericFnType|s are singleton types, {\em uniquely inhabited} by a single generic function.
-A consequence of this is that they are not useful as first-class values (i.e.
-they cannot be written into a collection).
-This is often valuable, particularly in parallel programming where compile-time specialization is valuable to avoid \textbf{performance} and \textbf{ease-of-use} issues that occur when using function pointers.
-
-Concrete functions, on the other hand, can be given a true function type (e.g.
-\verb|add5| could be compiled to a concrete function with type \verb|int| $\rightarrow$ \verb|int|) if targeting a backend that supports them, such as C99, or by using an integer-indexed jump table in OpenCL (we have not implemented this mechanism using Ace as of the time of writing, but do not anticipate difficulties in doing so).
-
-Type propagation via generic functions can be compared to template specialization in C++, where both the template headers (containing nested template parameters for function arguments) and specialization parameters at any call sites are inferred automatically from usage.
-This significantly simplifies a sophisticated feature of C++ and introduces it to OpenCL and C, which do not support templates.
-Other uses for C++ templates are subsumed by the metaprogramming features discussed in Section 5.
-% Is this really any more concise than using C + templates? Etc. relation to types. Conciseness.
+\todo{move this to related work??}Type assignment to generic functions is similar in some ways to template specialization in C++.
+In effect, both the template header and type parameters at call sites are being generated automatically by Ace.
+This significantly simplifies a sophisticated feature of C++ and introduces it to targets like OpenCL and C, which do not support templates.
+Other uses for C++ templates (and the preprocessor) are subsumed by the metaprogramming features discussed above\todo{cite template metaprogramming paper}.
 
@@
 \begin{codelisting}
-\lstinputlisting{listing8.py}
-\caption{[\texttt{listing\ref{py}.py}] A full OpenCL program using the \texttt{Ace.OpenCL} Python bindings, including data transfer to and from a device and direct invocation of a generic function, \texttt{map}, as a kernel without explicit compilation.}
-\label{py}
+\lstinputlisting{listing6.py}
+\caption{\texttt{[listing6.py]} A function demonstrating whole-function type inference when multiple values with differing types are assigned to a single identifier, \texttt{y}.}
+\label{inference}
 \end{codelisting}
-
-%\begin{codelisting}
-%\lstinputlisting{listing6.py}
-%\caption{\texttt{[listing6.py]} A function demonstrating whole-function type inference when multiple values with differing types are assigned to a single identifier, \texttt{y}.}
-%\label{inference}
-%\end{codelisting}
-
 On Line 5 in the generic \verb|map| function in Listing \ref{map}, the variable \verb|gid| is initialized with the result of calling the OpenCL primitive \verb|get_global_id|.
@@
 \subsection{Implicit Compilation and Execution}\label{compenv}\label{backend}\label{implicit}
+\begin{codelisting}
+\lstinputlisting{listing8.py}
+\caption{[\texttt{listing\ref{py}.py}] A full OpenCL program using the \texttt{Ace.OpenCL} Python bindings, including data transfer to and from a device and direct invocation of a generic function, \texttt{map}, as a kernel without explicit compilation.}
+\label{py}
+\end{codelisting}
 Professional end-users today generally use dynamically-typed high-level languages like MATLAB, Python, R or Perl for tasks that are not performance-sensitive, such as small-scale data analysis and plotting \cite{nguyen2010survey}.

### 2013-11-15 ace-pldi14/ace-pldi14.tex
pldi14 2.6 p1

@@
 To compile a generic function to a particular \emph{concrete function}, a type must be provided for each argument, and typechecking and translation must then succeed.
-Listing \ref{compscript} shows how to explicitly provide type assignments to \verb|map| using the subscript operator.
+Listing \ref{compscript} shows how to explicitly provide type assignments to \verb|map| using the subscript operator (implemented using Python's operator overloading mechanism).
 We attempt to do so three times in Listing \ref{compscript}.
@@
 This choice is merely an implementation detail that can be kept private to \verb|clx|, however.
-An Ace value of type \verb|clx.int2| (that is, an actual OpenCL vector) \emph{cannot} be used when a \verb|clx.Cplx(clx.int)| is expected (and attempting to do so will result in a static type error).% \verb|clx.Cplx| truly extends the type system, it is not a type alias.
+An Ace value of type \verb|clx.int2| (that is, an actual OpenCL vector) \emph{cannot} be used when a \verb|clx.Cplx(clx.int)| is expected (and attempting to do so will result in a static type error).
+\verb|clx.Cplx| truly extends the type system, it is not a type alias.
 
@@
 This also explains the error printed on line \ref{mapc}.3-\ref{mapc}.4: when this type was inadvertently assigned to the first argument \verb|input|, the indexing operation on line \ref{map}.6 resulted in an error.
-A generic function can only be \emph{statically} indexed by a list of types to turn it into a concrete function, not \emph{dynamically} indexed with a value of type \verb|clx.size_t| (the return type of the OpenCL primitive function \verb|get_global_id|).
+A generic function can only be \emph{statically} indexed by a list of types to turn it into a concrete function, not \emph{dynamically} indexed with a value of type \verb|clx.size_t| (the return type of the OpenCL primitive function \verb|get_global_id|, see below).
 
@@
 Concrete functions, on the other hand, can be given both a singleton type and a true function type.
-For example, \verb|listing2.scale[[clx.int, clx.int]]| could be given type \verb|ace.Arrow(clx.int, clx.int)|.
+For example, \verb|listing2.negate[[clx.int]]| could be given type \verb|ace.Arrow(clx.int, clx.int)|.
 The base determines how to convert the Ace arrow type to an arrow type in the target language (e.g.
-a function pointer for C99, or an integer that indexes into a jump table containing all functions in the program of that type).
+a function pointer for C99, or an integer that indexes into a jump table constructed from knowledge of available functions of the appropriate type in OpenCL).
 
@@
 
-\subsection{Whole-Function Type Inference}
-\begin{codelisting}
-\lstinputlisting{listing6.py}
-\caption{\texttt{[listing6.py]} A function demonstrating whole-function type inference when multiple values with differing types are assigned to a single identifier, \texttt{y}.}
-\label{inference}
-\end{codelisting}
-On Line 5 in the generic \verb|map| function in Listing \ref{map}, the variable \verb|gid| is initialized with the result of calling the OpenCL primitive \verb|get_global_id|.
+%\subsection{Within-Function Type Resolution}
+%\begin{codelisting}
+%\lstinputlisting{listing6.py}
+%\caption{\texttt{[listing6.py]} A function demonstrating whole-function type inference when multiple values with differing types are assigned to a single identifier, \texttt{y}.}
+%\label{inference}
+%\end{codelisting}
+%On line 5 in the generic \verb|map| function in Listing \ref{map}, the variable \verb|gid| is initialized with the result of calling the OpenCL primitive \verb|get_global_id|.
 The type for \verb|gid| is never given explicitly.
-This is a simple case of Ace's more general {\em whole-function type inference}.
-In this case, \verb|gid| will be inferred to have type \verb|size_t| because that is the return type of \verb|get_global_id| (as defined in the OpenCL specification, which the \verb|ace.OpenCL| module follows).
+This is a simple case of Ace's {\em within-function type resolution} strategy (we hesitate to call it \emph{type inference} because it does not, strictly speaking, take a constraint-solving approach).
+In this case, the type of \verb|gid| will resolve to \verb|size_t| because that is the return type of \verb|get_global_id| (as defined in the OpenCL specification, which the \verb|ace.OpenCL| module follows).
 The result can be observed on Lines 11 and 24 in Listing \ref{mapout}.
-
-Inference is not restricted within single assignments, as in the \verb|map| example, however.
+%
+%Inference is not restricted within single assignments, as in the \verb|map| example, however.
 Multiple assignments to the same identifier with values of differing types, or multiple return statements, can be combined if the types in each case are compatible with one another (e.g.
@@
 \subsection{Implicit Compilation and Execution}\label{compenv}\label{backend}\label{implicit}
+%Professional end-user programmers (e.g.
+scientists and engineers \cite{professional-end-users}) today generally use dynamically-typed high-level languages like MATLAB, Python, R or Perl for tasks that are not performance-sensitive, such as small-scale data analysis and plotting \cite{nguyen2010survey}.
+For portions of their analyses where the performance overhead of dynamic type checking and automatic memory management is too high, they will typically call into code written in a statically-typed, low-level language, most commonly C or Fortran, that uses low-level parallel abstractions like pthreads and MPI \cite{4222616,basili2008understanding}.
+Unfortunately, these low-level languages and abstractions are notoriously difficult to use and automatic verification is intractable in general.\todo{integrate this}
+
+A common workflow for \emph{professional end-user programmers} (e.g.
+scientists and engineers \cite{peus}) is to use a simple scripting language for orchestration, small-scale data analysis and visualization and call into a low-level language for performance-critical sections.
+Python is both designed for this style of use and widely adopted for such tasks \cite{sanner1999python,nguyen2010survey}.
+Developers can call into native functions using Python's foreign function interface (FFI), for example.
+A more recent trend is to generate and compile code without leaving Python, using a Python wrapper around a compiler.
+For example, \verb|weave| works with C and C++, and \verb|pycuda| and \verb|pyopencl| work with CUDA and OpenCL, respectively \cite{klockner2011pycuda}.
 \begin{codelisting}
@@
 \end{codelisting}
-Professional end-users today generally use dynamically-typed high-level languages like MATLAB, Python, R or Perl for tasks that are not performance-sensitive, such as small-scale data analysis and plotting \cite{nguyen2010survey}.
-For portions of their analyses where the performance overhead of dynamic type checking and automatic memory management is too high, they will typically call into code written in a statically-typed, low-level language, most commonly C or Fortran, that uses low-level parallel abstractions like pthreads and MPI \cite{4222616,basili2008understanding}.
-Unfortunately, these low-level languages and abstractions are notoriously difficult to use and automatic verification is intractable in general.\todo{integrate this}
-
-As discussed in the Introduction, a common workflow for professional end-users is to use a high-level scripting language for orchestration, small-scale data analysis and visualization and call into a low-level language for performance-critical sections.
-Python is designed for this style of use \cite{sanner1999python} and is widely used by professional end-users in HPC as a high-level scripting language.
-It features mature support for calling into code written in low-level languages.
-Developers can call into native libraries using its foreign function interface (FFI),  or by using a wrapper library like \verb|pycuda| for code compiled with CUDA \cite{klockner2011pycuda} or \verb|weave| for C and C++.
-
-Although C, C++ and CUDA's compilers are separate executables on the system, the OpenCL language was also designed for this workflow, in that it exposes the compiler and memory management directly as an API, called the \emph{host API}.
-The \verb|pyopencl| module exposes this API and supports basic interoperability with \verb|numpy|, the low-level linear algebra package for Python.
-With both \verb|pyopencl| and \verb|pycuda|, users generate OpenCL or CUDA source code as strings, compile it programmatically, then execute it using the run-time APIs that each library provides \cite{klockner2011pycuda}.
-This mode of use can be considered one where Python serves as an \emph{interactive compilation environment}.
+The OpenCL language was designed for this workflow, exposing a retargetable compiler and data management  routines as an API, called the \emph{host API} \cite{opencl11}.
+The \verb|pyopencl| library exposes this API and supports basic interoperability with \verb|numpy|, a package for safely manipulating contiguously-allocated numeric arrays in Python \cite{klockner2011pycuda}.
 

### 2013-11-15 ace-pldi14/ace-pldi14.tex
pldi14 section 2 first draft done (whew)

@@
 This also explains the error printed on line \ref{mapc}.3-\ref{mapc}.4: when this type was inadvertently assigned to the first argument \verb|input|, the indexing operation on line \ref{map}.6 resulted in an error.
-A generic function can only be \emph{statically} indexed by a list of types to turn it into a concrete function, not \emph{dynamically} indexed with a value of type \verb|clx.size_t| (the return type of the OpenCL primitive function \verb|get_global_id|, see below).
+A generic function can only be \emph{statically} indexed by a list of types to turn it into a concrete function, not \emph{dynamically} indexed with a value of type \verb|clx.size_t| (the return type of the OpenCL primitive function \verb|get_global_id|).
 
@@
 
-\subsection{Implicit Compilation and Execution}\label{compenv}\label{backend}\label{implicit}
+\subsection{Implicit Compilation and Interactive Execution}\label{compenv}\label{backend}\label{implicit}
 %Professional end-user programmers (e.g.
@@
 
-Ace supports a refinement to this workflow, as an alternative to the \verb|acec| compiler described above.
-In other words, \verb|acec| can generate source code for kernels but it does not specify how and from what language thet will be called.
-This mode of use allows Python to be the calling language.
+Ace supports a refinement to this workflow, as an alternative to the \verb|acec| compiler described above, for targets that have wrappers like this available, including \verb|clx.opencl|.
+Listing \ref{py} shows an example of this workflow where the user chooses a compute device (line \ref{py}.3),  constructs a \verb|numpy| array (line \ref{py}.5), transfers it to the device (line \ref{py}.6), allocates an empty equal-sized buffer for the result of the computation (line \ref{py}.7), launches the {generic} kernel \verb|map| from Listing \ref{map} with these device arrays as well as the function \verb|negate| from Listing \ref{metaprogramming} (line \ref{py}.9) choosing a number of threads equal to the number of elements in the input array (line \ref{py}.10), and transfers the result back into main memory to check that the device computed the expected result (line \ref{py}.12).
 
-At the time of writing, this is only supported with the OpenCL backend, but other backends can implement this feature as well by satisfying a simple interface specification.
-The OpenCL host API wraps and is a superset of the \verb|pyopencl|, adding a simpler type-aware API.
-Both generic functions and concrete functions can then be called like regular Python functions, with additional keyword arguments specifying the global and local size (i.e.
-the number of threads and how they are grouped).
-Device buffers can carry type information, unlike in the basic OpenCL host API, and this type information can be propagated from Python to Ace functions directly, as if the call had been within a kernel,  eliminating the \verb|compile| step that we have been using thus far and thus all mention of types.
+For developers experienced with the usual OpenCL or CUDA workflow, the fact that this can be accomplished in a total of 6 statements may be surprising.
+This simplicity is enabled largely due to  implicit uses of types throughout the code.
+First, \verb|numpy| keeps track of the type, shape and order of its arrays.
+The type of \verb|input|, for example, is \verb|numpy.double| by default, its shape is \verb|(1024,)| and its order is row-major by default.
+The \verb|pyopencl| library that our mechanism is built atop uses this information to automatically call the underlying OpenCL host API function for transferring byte arrays to the device without requiring the user to calculate the size.
+Our wrapper of \verb|pyopencl| retains this information in the Python representations of the device arrays, \verb|d_input| and \verb|d_output|.
+An Ace type parameterized can use this metadata to typecheck and generate code for operations like indexing in Ace functions\todo{rephrase}.
 
-An example of this for the generic \verb|map| function defined in Listing \ref{map} is shown in Listing \ref{py}, with the call itself on Lines 12-13. 
-The first two arguments to \verb|map| are OpenCL buffers, generated using a simplified wrapper to the \verb|pyopencl| APIs on Lines 9-10.
-This wrapper associates type information with each buffer, based on the type of the \verb|numpy| array, and this is used to implicitly compile \verb|map| as appropriate the first time it is called for any given combination of input types.
-Notice also that \verb|add5| is passed in directly.
+This allows us to call the generic function \verb|map| directly on these Python data structures (as well as the generic function \verb|negate|) without first requiring an explicit type assignment, like we needed when using \verb|acec| above.
+In other words, dynamic types and other metadata can propagate from Python data structures into an Ace generic function as static type information, in the same manner as it propagated \emph{between} generic functions in the previous section.
+In both cases, typechecking and translation of \verb|map| happens the first time a particular type assignment is encountered and cached for subsequent use.
+When called from Python, the generated OpenCL source code is also compiled for the device we selected using the OpenCL host API's compiler infrastructure, and cached.
 
-By way of comparison, the same program written using the OpenCL C API directly is an order of magnitude larger and significantly more difficult to read and understand.
-It does not support higher-order functions nor is there any way to write \verb|map| in a type-generic way.
-A full implementation of the logic of \verb|map| written using the \verb|pyopencl| bindings and metaprogramming techniques as described in \cite{klockner2011pycuda} is still twice as large and more difficult to comprehend than the code we have shown thus far.
-Not shown are several additional conveniences, such as delegated kernel sizing and \verb|In| and \verb|Out| constructs that can reduce the size and improve the clarity of this code further; due to a lack of space, the reader is referred to the language documentation.
+The same program written using the OpenCL C API directly is an order of magnitude longer and significantly more difficult to comprehend.
+OpenCL not support higher-order functions nor is there any way to write \verb|map| in a type-generic manner.
+If we instead use the \verb|pyopencl| library and apply the techniques described in \cite{klockner2011pycuda}, the program is still twice as large and less readable than this code.
+Both the \verb|map| and \verb|negate| functions must be explicitly specialized with the appropriate types using string manipulation techniques.
+Higher order functions are still not available, and must be simulated by string manipulation.
+That approach also does not permit the use any of the language extensions that Ace enables (in this case, the only interesting use of the extension mechanism was by the Ace type for \verb|numpy| arrays, which ensures that indexing respects ordering; see Sec.
+\ref{examples} for more interesting possibilities).\todo{would be aesthetically pleasing for a line to be shaved off so this ends up on the previous page}
+
+%Not shown are several additional conveniences, such as delegated kernel sizing and \verb|In| and \verb|Out| constructs that can reduce the size and improve the clarity of this code further; due to a lack of space, the reader is referred to the language documentation.
 

### 2013-11-15 ace-pldi14/listing8.py

@@ -1,12 +1,12 @@
 import listing1, listing2, examples.clx as clx, numpy
 
-clx.default_ctx = clx.Context.for_device(0, 0)
+clx.opencl.ctx = clx.Context.for_device(0, 0)
 
 input = numpy.ones((1024,))
 d_input = clx.to_device(input)
 d_output = clx.alloc(like=input)
 
-listing1.map(d_input, d_output, listing2.thresh_dbl, 
-  global_size=d_in.shape, local_size=(128,))
+listing1.map(d_input, d_output, listing2.negate, 
+  global_size=d_in.shape)
 
-assert (cl.from_device(d_out) == input * 2).all()
+assert (cl.from_device(d_out) == input * -1).all()

### 2013-11-15 ace-pldi14/ace-pldi14.tex
pldi14 section2 revision

@@
 \lstinputlisting{listing8.py}
-\caption{[\texttt{listing\ref{py}.py}] A full OpenCL program using the \texttt{Ace.OpenCL} Python bindings, including data transfer to and from a device and direct invocation of a generic function, \texttt{map}, as a kernel without explicit compilation.}
+\caption{[\texttt{listing\ref{py}.py}] A full OpenCL program using the \texttt{clx} Python bindings, including data transfer to and from a device and direct invocation of a generic function, \texttt{map}.}
 \label{py}
@@
 That approach also does not permit the use any of the language extensions that Ace enables (in this case, the only interesting use of the extension mechanism was by the Ace type for \verb|numpy| arrays, which ensures that indexing respects ordering; see Sec.
-\ref{examples} for more interesting possibilities).\todo{would be aesthetically pleasing for a line to be shaved off so this ends up on the previous page}
+\ref{examples} for more interesting possibilities).
 

### 2013-11-15 ace-pldi14/ace-pldi14.tex
pldi14 progress on sec 3

@@
 For developers experienced with the usual OpenCL or CUDA workflow, the fact that this can be accomplished in a total of 6 statements may be surprising.
-This simplicity is enabled largely due to  implicit uses of types throughout the code.
+This simplicity is enabled largely due to  implicit tracking of types throughout the code.
 First, \verb|numpy| keeps track of the type, shape and order of its arrays.
@@
 \section{Extensibility}\label{att}
+The core of Ace consists of about 1500 lines of Python code implementing its primary concepts: generic functions, concrete functions, active bases, active types and active targets.
+The latter three comprise Ace's extension mechanism.
+Extensions provide semantics to, and govern the compilation of, Ace functions, rather than logic in Ace's core.
+Indeed, the name ``Ace'' might itself be an acronym: it is an ``active compilation environment''.
+
+\subsection{Active Types}\label{atypes}
+Active types are the primary means for extending Ace, and motivate active bases, so we will describe them first.
+An active type, as mentioned above, is an instance of a class implementing the \verb|ace.ActiveType| interface.
+Listing \ref{pointers} shows an example of such a class: the \verb|clx.Ptr| class used in Listing \ref{compscript}, which implements the logic of OpenCL's pointer types.
+The class takes two arguments in its constructor: an address space and the target type of the pointer.
+OpenCL pointers support several operations: they can be indexed by an integer type to extract the value being pointed to; one can perform pointer arithmetic on them; and two pointers can be subtracted to produce a value of type \verb|ptrdiff_t| (a device-dependent integer type) if they are in the same address space \cite{opencl11}.
+
+In a compiler for a monolithic language, there would be a \emph{syntax-directed} protocol governing typechecking and translation.
+For example, in a functional language, one would declare datatypes that captured all forms of types and expressions, and the typechecker would perform exhaustive case analysis over the expression forms, implementing the logic described above in the cases for subscript (\verb|e1[e2]|) and addition (\verb|e1 + e2|).
+The visitor pattern typically used in object-oriented languages implements essentially the same protocol.
+This does not work for an extensible language: new cases need to be added, and the corresponding logic must be included.
+In Ace, new types reuse existing forms, so this requires that the logic inside each case be modifiable by extensions (in a safely composable way).
+
+Instead of taking a syntax-directed approach, the Ace compiler's typechecking and translation phases take a \emph{type-directed approach}.
+When encountering a compound expression (e.g.
+\verb|e1[e2]|), Ace defers control over typechecking and translation to a \emph{method} associated with the active type of a designated subexpression.
+The choice of subexpression is fixed by the language, motivated by a consideration of which expression plays a \emph{primary} role (for example, for subscripting, \verb|e1| is chosen).
+Collectively, these choices comprise the \emph{dispatch protocol} of Ace.
+Below are examples of these choices.
+Due to space constraints, we do not show the full protocol, consisting of a rule for each compound form in the grammar \cite{pythonast}.
+\begin{itemize}
+\item Responsibility over {\bf attribute access} (\texttt{e.attr}), {\bf subscripting} (\texttt{e[e1]}\todo{is it n-ary?}) and \textbf{calls} (\verb|e(e1, ..., en)|) and {\bf unary operations} (e.g.
+\verb|-e|) is handed to the type recursively assigned to \texttt{e}.
+\item Responsibility over {\bf binary operations} (e.g.
+\verb|e1 + e2|) is first handed to the type assigned to the left operand.
+If it indicates a type error, the type assigned to the right operand is handed responsibility, via a different method call.
+{Note that this operates like the corresponding rule in Python's \emph{dynamic} operator overloading mechanism; see Sec.
+\ref{related} for a discussion.}
+\item Responsibility over {\bf simple assignment statements}, is handed to the type of the variable on the left (which, as we will see below, is determined by the active base).
+\end{itemize}
+
 \subsection{Active Bases}\label{abases}
-\subsection{Active Types}\label{atypes}
+\begin{codelisting}
+\lstinputlisting{listing9.py}
+\caption{\texttt{[ace.OpenCL]} A portion of the implementation of OpenCL pointer types implementing subscripting logic using the Ace extension mechanism, AT\&T.}
+\label{pointers}
+\end{codelisting}
+Each generic function is associated with an active base, which is an object implementing the \verb|ace.ActiveBase| interface.
+The active base specifies the \emph{base semantics} of that function -- the semantics of statements and expressions that do not have a clear ``primary'' subexpression.
+
+For example, the type and translation of a numeric literal is determined by the active base.
+The active base \verb|clx.base|, for example, assigns integer literals the type \verb|clx.int32| while floating point literals have type \verb|clx.double|, to be consistent with the semantics of OpenCL and C99.
+The \verb|clx.base| object is an instance of \verb|clx.Base|.
+This class is written so that this choice can be made differently if desired.
+For example, the base \verb|clx.Base(flit_t=clx.float)| changes the base semantics so that floating point literals have type \verb|clx.float|.
+This is useful because some OpenCL devices do not support double-precision floating point, or impose a significant performance penalty for its use.
+For applications where this higher precision is not necessary, this is convenient.
+Similarly, a base where integer literals are represented using true integers, rather than 32-bit integers, might be useful in  applications where avoiding accidental overflows is more important than performance.
+True integers can be implemented as an active type (not shown).
+
+Indeed, to even use \verb|double| in OpenCL, an appropriate \verb|#pragma| must be inserted. The base detects uses of \verb|double| anywhere inside the function
+
 \subsection{Active Targets}\label{atargets}
@@
 \subsubsection{Active Typechecking}
-\begin{codelisting}
-\lstinputlisting{listing9.py}
-\caption{\texttt{[ace.OpenCL]} A portion of the implementation of OpenCL pointer types implementing subscripting logic using the Ace extension mechanism, AT\&T.}
-\label{pointers}
-\end{codelisting}
 
@@
 \subsubsection{Dispatch Protocol}
-Below are examples of the rules that comprise the Ace dispatch protocol.
-Due to space constraints, we do not list the entire dispatch protocol, which contains a rule for each possible syntactic form in the language.
-\begin{itemize}
-\item Responsibility over {\bf unary operations} like \verb|-x| is given to the type assigned to the operand, \verb|x|.
-\item Responsibility over {\bf binary operations} is first handed to the type assigned to the left operand.
-If it indicates that it does not understand the operation, the type assigned to the right operand is handed responsibility, via a different method call.
-{Note that this operates similarly to the Python run-time operator overloading protocol; see Section \ref{related}.}
-\item Responsibility over {\bf attribute access} (\texttt{e.attr}), {\bf subscript access} (\texttt{e[idx]}) and \textbf{calls} (\verb|e(e1, ..., en)|) is handed to the type assigned to \texttt{e}.
-\item In support of the type inference mechanism described in Section \ref{inference}, responsibility to resolve a type given multiple assignments can be taken by any of the types of the assignments, with priority given to later types.
-\end{itemize}
 

### 2013-11-15 ace-pldi14/ace-pldi14.tex
Made table beautiful

--- 2013-11-15 ace-pldi14/ace-pldi14.tex
@@
 % authoryear    To obtain author/year citation style instead of numeric.
+
 \usepackage{listings}
@@
 
+% Heather-added packages for the fancy table
+\usepackage{longtable}
+\usepackage{booktabs}
+\usepackage{pdflscape}
+\usepackage{colortbl}%
+  \newcommand{\myrowcolour}{\rowcolor[gray]{0.925}}
+\usepackage{wasysym}
+
     \makeatletter
-%
+
 % \btIfInRange{number}{range list}{TRUE}{FALSE}
@@
 To compile a generic function to a particular \emph{concrete function}, a type must be provided for each argument, and typechecking and translation must then succeed.
-Listing \ref{compscript} shows how to explicitly provide type assignments to \verb|map| using the subscript operator (implemented using Python's operator overloading mechanism).
+Listing \ref{compscript} shows how to explicitly provide type assignments to \verb|map| using the subscript operator.
 We attempt to do so three times in Listing \ref{compscript}.
@@
 This choice is merely an implementation detail that can be kept private to \verb|clx|, however.
-An Ace value of type \verb|clx.int2| (that is, an actual OpenCL vector) \emph{cannot} be used when a \verb|clx.Cplx(clx.int)| is expected (and attempting to do so will result in a static type error).
-\verb|clx.Cplx| truly extends the type system, it is not a type alias.
+An Ace value of type \verb|clx.int2| (that is, an actual OpenCL vector) \emph{cannot} be used when a \verb|clx.Cplx(clx.int)| is expected (and attempting to do so will result in a static type error).% \verb|clx.Cplx| truly extends the type system, it is not a type alias.
 
@@
 Generic functions, according to Sec.
-\ref{genfn}, have not yet had a type assigned to them, so it may seem perplexing that we are nevertheless extracting a type from it.
-Although a conventional arrow type cannot be assigned to \verb|negate|, we can give it a \emph{singleton type}: a type that simply means ``this expression is the \emph{particular} generic function \verb|negate|''.
-This type could also have been explicitly written as \verb|ace.GenericFnType(listing2.negate)|.
-During typechecking and translation of \verb|map_neg_f32| and \verb|map_neg_ci32|, the call to \verb|f| on line \ref{map}.6 operates by using the types of the provided arguments to compile the generic function that inhabits the singleton type of \verb|f| (\verb|negate| in both of these cases) to a concrete function.
-This is why there are two versions of \verb|negate| in the output in Listing \ref{mapout}.
-In other words, types \emph{propagate} into generic functions -- we didn't need to compile \verb|negate| explicitly.
-This also explains the error printed on line \ref{mapc}.3-\ref{mapc}.4: when this type was inadvertently assigned to the first argument \verb|input|, the indexing operation on line \ref{map}.6 resulted in an error.
-A generic function can only be \emph{statically} indexed by a list of types to turn it into a concrete function, not \emph{dynamically} indexed with a value of type \verb|clx.size_t| (the return type of the OpenCL primitive function \verb|get_global_id|).
+\ref{genfn}, have not yet had a type assigned to them, so it may seem perplexing that we are nevertheless assigning a type to it.
 
-In effect, this scheme enables higher-order functions even when targeting languages, like OpenCL, that have no support for higher-order functions (OpenCL, unlike C99, does not support function pointers).
-Interestingly, because they have a singleton type, they are higher-order but not first-class functions.
-That is, the type system would prevent you from creating a heterogeneous list of generic functions.
-Concrete functions, on the other hand, can be given both a singleton type and a true function type.
-For example, \verb|listing2.negate[[clx.int]]| could be given type \verb|ace.Arrow(clx.int, clx.int)|.
-The base determines how to convert the Ace arrow type to an arrow type in the target language (e.g.
-a function pointer for C99, or an integer that indexes into a jump table constructed from knowledge of available functions of the appropriate type in OpenCL).
+This notion of types as metalanguage objects is key to the Ace compilation model and also enables other mechanisms that we will discuss in subsequent sections.
 
-\todo{move this to related work??}Type assignment to generic functions is similar in some ways to template specialization in C++.
-In effect, both the template header and type parameters at call sites are being generated automatically by Ace.
-This significantly simplifies a sophisticated feature of C++ and introduces it to targets like OpenCL and C, which do not support templates.
-Other uses for C++ templates (and the preprocessor) are subsumed by the metaprogramming features discussed above\todo{cite template metaprogramming paper}.
+\subsection{Type Propagation}
+The type assigned to the third argument, \verb|f|, on both Lines 4.9 and 4.10, is \verb|add5.ace_type|.
+The \verb|ace_type| attribute of a generic function is an instance of \verb|ace.GenericFnType|, the type of Ace generic functions.
+Ace generic functions are compiled to concrete functions automatically at all internal call sites.
+That is, when the compiler encounters the call to \verb|f| inside \verb|map| when compiling \verb|map_add5_double|, it compiles a version of \verb|add5| specialized to the \verb|double| type (seen on Line 5.3), and similarly when compiling \verb|map_add5_int| (on Line 5.16, automatically given a unique name to avoid conflicts).
+This mechanism is called {\em type propagation}.
+We did not need to use \verb|add5.compile(double)| before compiling \verb|map_add5_dbl| because only functions that are never called in the process of compiling other functions in a module need type information explicitly provided, supporting \textbf{ease-of-use} by increasing conciseness.
 
-%\subsection{Within-Function Type Resolution}
+In effect, this scheme allows for a form of higher-order functional programming even when targeting languages, like OpenCL, that have no support for higher-order functions (OpenCL, unlike C99, does not support function pointers).
+This works because the \verb|ace.GenericFnType| for one function, such as \verb|add5|, is not equal to  the \verb|ace.GenericFnType| for a superficially similar function, such as \verb|add6| (defined as one would expect).
+To put it in type theoretic terms, \verb|ace.GenericFnType|s are singleton types, {\em uniquely inhabited} by a single generic function.
+A consequence of this is that they are not useful as first-class values (i.e.
+they cannot be written into a collection).
+This is often valuable, particularly in parallel programming where compile-time specialization is valuable to avoid \textbf{performance} and \textbf{ease-of-use} issues that occur when using function pointers.
+
+Concrete functions, on the other hand, can be given a true function type (e.g.
+\verb|add5| could be compiled to a concrete function with type \verb|int| $\rightarrow$ \verb|int|) if targeting a backend that supports them, such as C99, or by using an integer-indexed jump table in OpenCL (we have not implemented this mechanism using Ace as of the time of writing, but do not anticipate difficulties in doing so).
+
+Type propagation via generic functions can be compared to template specialization in C++, where both the template headers (containing nested template parameters for function arguments) and specialization parameters at any call sites are inferred automatically from usage.
+This significantly simplifies a sophisticated feature of C++ and introduces it to OpenCL and C, which do not support templates.
+Other uses for C++ templates are subsumed by the metaprogramming features discussed in Section 5.
+% Is this really any more concise than using C + templates? Etc. relation to types. Conciseness.
+
+\subsection{Whole-Function Type Inference}
+\begin{codelisting}
+\lstinputlisting{listing8.py}
+\caption{[\texttt{listing\ref{py}.py}] A full OpenCL program using the \texttt{Ace.OpenCL} Python bindings, including data transfer to and from a device and direct invocation of a generic function, \texttt{map}, as a kernel without explicit compilation.}
+\label{py}
+\end{codelisting}
+
 %\begin{codelisting}
@@
 %\end{codelisting}
-%On line 5 in the generic \verb|map| function in Listing \ref{map}, the variable \verb|gid| is initialized with the result of calling the OpenCL primitive \verb|get_global_id|.
+
+On Line 5 in the generic \verb|map| function in Listing \ref{map}, the variable \verb|gid| is initialized with the result of calling the OpenCL primitive \verb|get_global_id|.
 The type for \verb|gid| is never given explicitly.
-This is a simple case of Ace's {\em within-function type resolution} strategy (we hesitate to call it \emph{type inference} because it does not, strictly speaking, take a constraint-solving approach).
-In this case, the type of \verb|gid| will resolve to \verb|size_t| because that is the return type of \verb|get_global_id| (as defined in the OpenCL specification, which the \verb|ace.OpenCL| module follows).
+This is a simple case of Ace's more general {\em whole-function type inference}.
+In this case, \verb|gid| will be inferred to have type \verb|size_t| because that is the return type of \verb|get_global_id| (as defined in the OpenCL specification, which the \verb|ace.OpenCL| module follows).
 The result can be observed on Lines 11 and 24 in Listing \ref{mapout}.
-%
-%Inference is not restricted within single assignments, as in the \verb|map| example, however.
+
+Inference is not restricted within single assignments, as in the \verb|map| example, however.
 Multiple assignments to the same identifier with values of differing types, or multiple return statements, can be combined if the types in each case are compatible with one another (e.g.
@@
 
-\subsection{Implicit Compilation and Interactive Execution}\label{compenv}\label{backend}\label{implicit}
-%Professional end-user programmers (e.g.
-scientists and engineers \cite{professional-end-users}) today generally use dynamically-typed high-level languages like MATLAB, Python, R or Perl for tasks that are not performance-sensitive, such as small-scale data analysis and plotting \cite{nguyen2010survey}.
+\subsection{Implicit Compilation and Execution}\label{compenv}\label{backend}\label{implicit}
+Professional end-users today generally use dynamically-typed high-level languages like MATLAB, Python, R or Perl for tasks that are not performance-sensitive, such as small-scale data analysis and plotting \cite{nguyen2010survey}.
 For portions of their analyses where the performance overhead of dynamic type checking and automatic memory management is too high, they will typically call into code written in a statically-typed, low-level language, most commonly C or Fortran, that uses low-level parallel abstractions like pthreads and MPI \cite{4222616,basili2008understanding}.
@@
 
-A common workflow for \emph{professional end-user programmers} (e.g.
-scientists and engineers \cite{peus}) is to use a simple scripting language for orchestration, small-scale data analysis and visualization and call into a low-level language for performance-critical sections.
-Python is both designed for this style of use and widely adopted for such tasks \cite{sanner1999python,nguyen2010survey}.
-Developers can call into native functions using Python's foreign function interface (FFI), for example.
-A more recent trend is to generate and compile code without leaving Python, using a Python wrapper around a compiler.
-For example, \verb|weave| works with C and C++, and \verb|pycuda| and \verb|pyopencl| work with CUDA and OpenCL, respectively \cite{klockner2011pycuda}.
-\begin{codelisting}
-\lstinputlisting{listing8.py}
-\caption{[\texttt{listing\ref{py}.py}] A full OpenCL program using the \texttt{Ace.OpenCL} Python bindings, including data transfer to and from a device and direct invocation of a generic function, \texttt{map}, as a kernel without explicit compilation.}
-\label{py}
-\end{codelisting}
-The OpenCL language was designed for this workflow, exposing a retargetable compiler and data management  routines as an API, called the \emph{host API} \cite{opencl11}.
-The \verb|pyopencl| library exposes this API and supports basic interoperability with \verb|numpy|, a package for safely manipulating contiguously-allocated numeric arrays in Python \cite{klockner2011pycuda}.
+As discussed in the Introduction, a common workflow for professional end-users is to use a high-level scripting language for orchestration, small-scale data analysis and visualization and call into a low-level language for performance-critical sections.
+Python is designed for this style of use \cite{sanner1999python} and is widely used by professional end-users in HPC as a high-level scripting language.
+It features mature support for calling into code written in low-level languages.
+Developers can call into native libraries using its foreign function interface (FFI),  or by using a wrapper library like \verb|pycuda| for code compiled with CUDA \cite{klockner2011pycuda} or \verb|weave| for C and C++.
 
-Ace supports a refinement to this workflow, as an alternative to the \verb|acec| compiler described above, for targets that have wrappers like this available, including \verb|clx.opencl|.
-Listing \ref{py} shows an example of this workflow where the user chooses a compute device (line \ref{py}.3),  constructs a \verb|numpy| array (line \ref{py}.5), transfers it to the device (line \ref{py}.6), allocates an empty equal-sized buffer for the result of the computation (line \ref{py}.7), launches the {generic} kernel \verb|map| from Listing \ref{map} with these device arrays as well as the function \verb|negate| from Listing \ref{metaprogramming} (line \ref{py}.9) choosing a number of threads equal to the number of elements in the input array (line \ref{py}.10), and transfers the result back into main memory to check that the device computed the expected result (line \ref{py}.12).
+Although C, C++ and CUDA's compilers are separate executables on the system, the OpenCL language was also designed for this workflow, in that it exposes the compiler and memory management directly as an API, called the \emph{host API}.
+The \verb|pyopencl| module exposes this API and supports basic interoperability with \verb|numpy|, the low-level linear algebra package for Python.
+With both \verb|pyopencl| and \verb|pycuda|, users generate OpenCL or CUDA source code as strings, compile it programmatically, then execute it using the run-time APIs that each library provides \cite{klockner2011pycuda}.
+This mode of use can be considered one where Python serves as an \emph{interactive compilation environment}.
 
-For developers experienced with the usual OpenCL or CUDA workflow, the fact that this can be accomplished in a total of 6 statements may be surprising.
-This simplicity is enabled largely due to  implicit uses of types throughout the code.
-First, \verb|numpy| keeps track of the type, shape and order of its arrays.
-The type of \verb|input|, for example, is \verb|numpy.double| by default, its shape is \verb|(1024,)| and its order is row-major by default.
-The \verb|pyopencl| library that our mechanism is built atop uses this information to automatically call the underlying OpenCL host API function for transferring byte arrays to the device without requiring the user to calculate the size.
-Our wrapper of \verb|pyopencl| retains this information in the Python representations of the device arrays, \verb|d_input| and \verb|d_output|.
-An Ace type parameterized can use this metadata to typecheck and generate code for operations like indexing in Ace functions\todo{rephrase}.
+Ace supports a refinement to this workflow, as an alternative to the \verb|acec| compiler described above.
+In other words, \verb|acec| can generate source code for kernels but it does not specify how and from what language thet will be called.
+This mode of use allows Python to be the calling language.
 
-This allows us to call the generic function \verb|map| directly on these Python data structures (as well as the generic function \verb|negate|) without first requiring an explicit type assignment, like we needed when using \verb|acec| above.
-In other words, dynamic types and other metadata can propagate from Python data structures into an Ace generic function as static type information, in the same manner as it propagated \emph{between} generic functions in the previous section.
-In both cases, typechecking and translation of \verb|map| happens the first time a particular type assignment is encountered and cached for subsequent use.
-When called from Python, the generated OpenCL source code is also compiled for the device we selected using the OpenCL host API's compiler infrastructure, and cached.
+At the time of writing, this is only supported with the OpenCL backend, but other backends can implement this feature as well by satisfying a simple interface specification.
+The OpenCL host API wraps and is a superset of the \verb|pyopencl|, adding a simpler type-aware API.
+Both generic functions and concrete functions can then be called like regular Python functions, with additional keyword arguments specifying the global and local size (i.e.
+the number of threads and how they are grouped).
+Device buffers can carry type information, unlike in the basic OpenCL host API, and this type information can be propagated from Python to Ace functions directly, as if the call had been within a kernel,  eliminating the \verb|compile| step that we have been using thus far and thus all mention of types.
 
-The same program written using the OpenCL C API directly is an order of magnitude longer and significantly more difficult to comprehend.
-OpenCL not support higher-order functions nor is there any way to write \verb|map| in a type-generic manner.
-If we instead use the \verb|pyopencl| library and apply the techniques described in \cite{klockner2011pycuda}, the program is still twice as large and less readable than this code.
-Both the \verb|map| and \verb|negate| functions must be explicitly specialized with the appropriate types using string manipulation techniques.
-Higher order functions are still not available, and must be simulated by string manipulation.
-That approach also does not permit the use any of the language extensions that Ace enables (in this case, the only interesting use of the extension mechanism was by the Ace type for \verb|numpy| arrays, which ensures that indexing respects ordering; see Sec.
-\ref{examples} for more interesting possibilities).\todo{would be aesthetically pleasing for a line to be shaved off so this ends up on the previous page}
+An example of this for the generic \verb|map| function defined in Listing \ref{map} is shown in Listing \ref{py}, with the call itself on Lines 12-13. 
+The first two arguments to \verb|map| are OpenCL buffers, generated using a simplified wrapper to the \verb|pyopencl| APIs on Lines 9-10.
+This wrapper associates type information with each buffer, based on the type of the \verb|numpy| array, and this is used to implicitly compile \verb|map| as appropriate the first time it is called for any given combination of input types.
+Notice also that \verb|add5| is passed in directly.
 
-%Not shown are several additional conveniences, such as delegated kernel sizing and \verb|In| and \verb|Out| constructs that can reduce the size and improve the clarity of this code further; due to a lack of space, the reader is referred to the language documentation.
+By way of comparison, the same program written using the OpenCL C API directly is an order of magnitude larger and significantly more difficult to read and understand.
+It does not support higher-order functions nor is there any way to write \verb|map| in a type-generic way.
+A full implementation of the logic of \verb|map| written using the \verb|pyopencl| bindings and metaprogramming techniques as described in \cite{klockner2011pycuda} is still twice as large and more difficult to comprehend than the code we have shown thus far.
+Not shown are several additional conveniences, such as delegated kernel sizing and \verb|In| and \verb|Out| constructs that can reduce the size and improve the clarity of this code further; due to a lack of space, the reader is referred to the language documentation.
 
@@
 
-\section{Related Work}\label{related}
+%\section{Related Work}\label{related}
+%\begin{figure*}
+%\begin{tabular}[t]{ l | l | c c c c c }
+%\textbf{Approach} & \textbf{Examples} & Library & {Extensible} & Extensible & Composable & Alternative \\
+%\hline
+%Active Types & Ace & x & ~ & x & x & x\\
+%Desugaring & SugarJ \cite{erdweg2011sugarj}, Sugar* & ~ & x & ~ & ~ & ~\\
+%Rule Injection & Qi, Typed Racket \cite{TypedScheme2008}/Clojure, A? & 1+2 & ~ & x & ~ & ~\\
+%Static metaprogramming & Scala macros \cite{ScalaMacros2013}, OJ \cite{OpenJava2000}, OC++ \cite{OpenCpp1995}, MorphJ \cite{MorphJ2011} & ~ & ~ & ~ & x & ~\\
+%Cross-compilation & Delite \cite{Delite2011} & x & ~ & ~ & ~ & x\\
+%EDSL frameworks & ? & x & x & x & ~ & ~\\
+%Type-specific literals & Wyvern & ~ & x & ~ & x & ~\\
+%\end{tabular}
+%\caption{Comparison to related approaches}
+%\end{figure*}
+
+\parindent0pt
+%
 \begin{figure*}
-\begin{tabular}[t]{ l | l | c c c c c }
-\textbf{Approach} & \textbf{Examples} & Library & {Extensible} & Extensible & Composable & Alternative \\
-\hline
-Active Types & Ace & x & ~ & x & x & x\\
-Desugaring & SugarJ, Sugar* & ~ & x & ~ & ~ & ~\\
-Rule Injection & Qi, Typed Racket/Clojure, A? & 1+2 & ~ & x & ~ & ~\\
-Static metaprogramming & Scala macros, OJ, OC++ & ~ & ~ & ~ & x & ~\\
-Cross-compilation & Delite & x & ~ & ~ & ~ & x\\
-EDSL frameworks & ? & x & x & x & ~ & ~\\
-Type-specific literals & Wyvern & ~ & x & ~ & x & ~\\
-\end{tabular}
+\onecolumn
+\begin{longtable}{l l@{}l c@{}c c@{}c c@{}c c@{}c c@{}c}
+\toprule%
+
+&&  &&  && {\bfseries Extensible} && {\bfseries Extensible} &&  && \\
+
+{\bfseries Approach} && {\bfseries Examples} && {\bfseries Library} && {\bfseries Syntax} && {\bfseries Type System} && {\bfseries Composable} && {\bfseries Alternative}  \\
+
+\cmidrule(l){1-1} \cmidrule(l){2-3} \cmidrule(l){4-5} \cmidrule(l){6-7} \cmidrule(l){8-9} \cmidrule(l){10-11} \cmidrule(l){12-13}
+
+\endhead
+
+Active Types && Ace && \CIRCLE && \Circle && \CIRCLE && \CIRCLE && \CIRCLE \\
+
+\myrowcolour%
+Desugaring && SugarJ \cite{erdweg2011sugarj}, Sugar* && \Circle && \CIRCLE && \Circle && \Circle && \Circle \\
+
+Rule Injection && Qi, Typed Racket \cite{TypedScheme2008}/Clojure, A? && 1+2 && \Circle && \CIRCLE && \Circle && \Circle \\
+
+\myrowcolour%
+Static Metaprogramming && Scala macros \cite{ScalaMacros2013}, OJ \cite{OpenJava2000}, && \Circle && \Circle && \Circle && \CIRCLE && \Circle \\
+
+\myrowcolour
+ && OC++ \cite{OpenCpp1995}, MorphJ \cite{MorphJ2011} && && && && && \\
+
+Cross-Compilation && Delite \cite{Delite2011} && \CIRCLE && \Circle && \Circle && \Circle && \CIRCLE \\
+
+\myrowcolour%
+EDSL Frameworks && ? && \CIRCLE && \CIRCLE && \CIRCLE && \Circle && \Circle \\
+
+Type-Specific Literals && Wyvern && \Circle && \CIRCLE && \Circle && \CIRCLE && \Circle \\
+
+\bottomrule
+
+\end{longtable}
 \caption{Comparison to related approaches}
+\twocolumn
 \end{figure*}
+
 \subsection{Active Libraries}

### 2013-11-15 ace-pldi14/ace-pldi14.tex
pldi14 sectionized sec 3

--- 2013-11-15 ace-pldi14/ace-pldi14.tex
@@
 When encountering a compound expression (e.g.
-\verb|e1[e2]|), Ace defers control over typechecking and translation to a \emph{method} associated with the active type of a designated subexpression.
-The choice of subexpression is fixed by the language, motivated by a consideration of which expression plays a \emph{primary} role (for example, for subscripting, \verb|e1| is chosen).
+\verb|e1[e2]|), Ace defers control over typechecking and translation to the active type of a designated subexpression.
+The choice of subexpression is fixed by the language, motivated by considerations of Python's semantics for the same form.
 Collectively, these choices comprise the \emph{dispatch protocol} of Ace.
@@
 \item Responsibility over {\bf simple assignment statements}, is handed to the type of the variable on the left (which, as we will see below, is determined by the active base).
+If this type does not provide special assignment semantics, the base handles it.
+%Destructuring assignment is also supported by a somewhat more complex protocol.\todo{clarify}
 \end{itemize}
 
-\subsection{Active Bases}\label{abases}
+\subsubsection{Active Typechecking}
 \begin{codelisting}
@@
 \end{codelisting}
+When the compiler encounters an expression, it must first verify its validity by either assigning it a type or raising a meaningful type error.
+Rather than defining fixed logic for this, Ace defers control to the {type} of the \emph{primary operand} according to a {\em dispatch protocol}.
+Because types are metalanguage instances of user-defined types derived from \verb|ace.Type| (cf.
+Section 2.5), dispatch corresponds to calling a method of this base class.
+
+Let us again consider the data-parallel map example of Section 2.
+When \verb|map| is compiled on Line 9 of Listing \ref{compscript}, the first two argument types are \verb|gptr(double)|.
+This type is an instance of the user-defined \verb|OpenCL.PtrType| which inherits from \verb|ace.Type|.
+When the compiler encounters the expression \verb|input[gid]| on Line 6, the dispatch protocol is to defer control to the type of \verb|input| by invoking the method named \verb|resolve_Subscript|.
+
+The relevant portion of \verb|OpenCL.PtrType| as well as \verb|gptr| is shown in Listing \ref{pointers}.
+The \verb|resolve_Subscript| method on Line 8 receives a context and the syntax tree of the node being considered.
+The context contains information about variables in scope and other contextual information, and also exposes a method, \verb|resolve_type|, that this method uses to recursively resolve the types of other subexpressions, here the slice which will be \verb|gid|, as needed.
+The context will have that \verb|gid| is the machine-dependent integer type \verb|size_t| as discussed in Section 2.
+On Line 10, it confirms that this type is an instance of an integer type and thus, it assigns the whole expression, \verb|input[gid]|, the target type of the pointer, \verb|double|.
+Had the function attempted to index \verb|input| using a non-integer expression, the method would take the other branch of the conditional and raise a type error, with a custom error message, on Line 13.
+We note that error messages are an important component of \textbf{ease-of-use} \cite{marceau2011measuring}.
+Indeed, a widely-reported frustration with C++ is that it produces overly verbose and cryptic error messages, particularly when templates are used in clever ways.
+
+\subsubsection{Active Translation}
+Once typechecking a method is complete, the compiler must subsequently translate each Ace source expression into an expression in the target language.
+This has been OpenCL in the examples thus far, but we describe a generalization of this in the next section.
+
+It does so by again applying the dispatch protocol to call a method of the form \verb|translate_|$X$, where $X$ is the syntactic form of the expression.
+This method is responsible for returning a copy of the expression's ast node with an additional attribute, \verb|code|, containing the source code of the translation, represented here as a string though it may also be represented in a structured manner.
+In our example, it is simply a direct translation to the corresponding OpenCL attribute access (Line 20), using the recursively-determined translations of the operands (Lines 16-17).
+More sophisticated abstractions may insert arbitrarily complex statements and expressions during this phase.
+The context also provides some support for non-local insertions, such as new top-level type declarations, imports and helper code (not shown).
+
+\subsection{Active Bases}\label{abases}
 Each generic function is associated with an active base, which is an object implementing the \verb|ace.ActiveBase| interface.
@@
 Thus, as described so far, Ace has affirmatively resolved the question it was originally conceived to address, discussed in Section 2.1.
-
-\subsection{Monolithic vs. Extensible Languages}
-OpenCL is not necessarily the best tool for every job in high-performance computing.
-Indeed, HPC is an area where designing a set of primitives that satisfy all users has been particularly challenging, and it appears unlikely that a broad consensus will emerge given the variety of architectures, applications, scales and user communities that it serves, and the number of seemingly promising abstractions that emerge continuously targeting various subsets of this problem space.
-
-It is therefore a concern that most programming languages are {\em monolithic} -- a collection of primitives are given first-class treatment by the language implementation, and users can only creatively combine them to implement algorithms and abstractions of their design.
-Although highly-expressive general-purpose mechanisms have been developed (such as object systems or algebraic datatypes), these may not suffice when researchers or domain experts wish to evolve aspects of the type system, exert control over the representation of data, introduce specialized run-time mechanisms, or if defining an abstraction in terms of existing mechanisms is unnatural or verbose (in summary, to push the boundaries of \textbf{verifiability}, \textbf{performance} and \textbf{ease-of-use}).
-In these situations, it would be desirable to have the ability to modularly extend existing systems (\textbf{continuity}) with new compile-time logic (\textbf{extensibility}) and be assured that such extensions will never interfere with one another when used in the same program (\textbf{interoperability}).
-
-%Because Ace libraries can contain compile-time logic, written in the metalanguage as described in the previous section, these primitive definitions can be distributed as modules, rather than as extensions to particular compilers or domain-specific languages.
-\subsection{Active Typechecking and Translation}
-To achieve these criteria, Ace has been designed around a minimal, {extensible} core.
-Users introduce the compile-time logic associated with primitive types and operations from \emph{within libraries}, as opposed to by some language-external mechanism such as a domain-specific language framework or extensible compiler (see Section \ref{related}).
-
-The two phases of compilation that can be controlled by users are together referred to as \emph{active typechecking and translation (AT\&T)}.
-They are invoked the first time the \verb|compile| method of a generic function is called, or when type propagation occurs, with a particular type assignment (a cached concrete function is returned subsequently).
-
-\subsubsection{Active Typechecking}
-
-%We now explain how new primitive types and operators can be defined in Ace. 
-When the compiler encounters an expression, it must first verify its validity by either assigning it a type or raising a meaningful type error.
-Rather than defining fixed logic for this, Ace defers control to the {type} of the \emph{primary operand} according to a {\em dispatch protocol}.
-Because types are metalanguage instances of user-defined types derived from \verb|ace.Type| (cf.
-Section 2.5), dispatch corresponds to calling a method of this base class.
-
-Let us again consider the data-parallel map example of Section 2.
-When \verb|map| is compiled on Line 9 of Listing \ref{compscript}, the first two argument types are \verb|gptr(double)|.
-This type is an instance of the user-defined \verb|OpenCL.PtrType| which inherits from \verb|ace.Type|.
-When the compiler encounters the expression \verb|input[gid]| on Line 6, the dispatch protocol is to defer control to the type of \verb|input| by invoking the method named \verb|resolve_Subscript|.
-
-The relevant portion of \verb|OpenCL.PtrType| as well as \verb|gptr| is shown in Listing \ref{pointers}.
-The \verb|resolve_Subscript| method on Line 8 receives a context and the syntax tree of the node being considered.
-The context contains information about variables in scope and other contextual information, and also exposes a method, \verb|resolve_type|, that this method uses to recursively resolve the types of other subexpressions, here the slice which will be \verb|gid|, as needed.
-The context will have that \verb|gid| is the machine-dependent integer type \verb|size_t| as discussed in Section 2.
-On Line 10, it confirms that this type is an instance of an integer type and thus, it assigns the whole expression, \verb|input[gid]|, the target type of the pointer, \verb|double|.
-Had the function attempted to index \verb|input| using a non-integer expression, the method would take the other branch of the conditional and raise a type error, with a custom error message, on Line 13.
-We note that error messages are an important component of \textbf{ease-of-use} \cite{marceau2011measuring}.
-Indeed, a widely-reported frustration with C++ is that it produces overly verbose and cryptic error messages, particularly when templates are used in clever ways.
-
-\subsubsection{Dispatch Protocol}
-
-\subsubsection{Active Translation}
-Once typechecking a method is complete, the compiler must subsequently translate each Ace source expression into an expression in the target language.
-This has been OpenCL in the examples thus far, but we describe a generalization of this in the next section.
-
-It does so by again applying the dispatch protocol to call a method of the form \verb|translate_|$X$, where $X$ is the syntactic form of the expression.
-This method is responsible for returning a copy of the expression's ast node with an additional attribute, \verb|code|, containing the source code of the translation, represented here as a string though it may also be represented in a structured manner.
-In our example, it is simply a direct translation to the corresponding OpenCL attribute access (Line 20), using the recursively-determined translations of the operands (Lines 16-17).
-More sophisticated abstractions may insert arbitrarily complex statements and expressions during this phase.
-The context also provides some support for non-local insertions, such as new top-level type declarations, imports and helper code (not shown).
-
-\subsubsection{User-Defined Backends}\label{backends}
-Thus far, we have discussed using OpenCL itself as a backend for our implementation of the OpenCL primitives.
-This backend is called \verb|OpenCL| in the \verb|ace.OpenCL| module.
-Ace supports the definition of new backends in a manner similar to the introduction of new types, by extending the \verb|ace.Backend| base class.
-Backends are specified for a function by using the decorator form \verb|@backend.fn|, as can be seen in the previous examples.
-Backends are responsible for some aspects of the grammar that do not admit simple dispatch to the type of a subterm, such as number and string literals or statement forms like \verb|while| (though \verb|for| is handled by an iterator-like protocol, not shown).
-
-In addition to the OpenCL backend, preliminary C99 and CUDA backends are available (with the caveat that they have not been as fully developed or tested as of this writing.) This allows us to use the OpenCL kernel language, which offers a simplified variant of C99, without relying on the full OpenCL stack, which may not be well-supported by  vendors with proprietary solutions such as CUDA.
-Backends not based on the C family are also possible, but we leave such developments for future work.
-During the translation phase, types can access the backend via the context and emit different code for different supported backends (not shown above due to space considerations).
 
@@
 
+It is therefore a concern that most programming languages are {\em monolithic} -- a collection of primitives are given first-class treatment by the language implementation, and users can only creatively combine them to implement algorithms and abstractions of their design.
+Although highly-expressive general-purpose mechanisms have been developed (such as object systems or algebraic datatypes), these may not suffice when researchers or domain experts wish to evolve aspects of the type system, exert control over the representation of data, introduce specialized run-time mechanisms, or if defining an abstraction in terms of existing mechanisms is unnatural or verbose (in summary, to push the boundaries of \textbf{verifiability}, \textbf{performance} and \textbf{ease-of-use}).
+In these situations, it would be desirable to have the ability to modularly extend existing systems (\textbf{continuity}) with new compile-time logic (\textbf{extensibility}) and be assured that such extensions will never interfere with one another when used in the same program (\textbf{interoperability}).
+
+OpenCL is not necessarily the best tool for every job in high-performance computing.
+Indeed, HPC is an area where designing a set of primitives that satisfy all users has been particularly challenging, and it appears unlikely that a broad consensus will emerge given the variety of architectures, applications, scales and user communities that it serves, and the number of seemingly promising abstractions that emerge continuously targeting various subsets of this problem space.
+
 It is legitimate to ask, however, why dynamically-typed languages are so widely-used in HPC.

### 2013-11-15 ace-pldi14/ace-pldi14.tex
pldi14 sectionized sec 4 roughly

@@
 \section{Examples}\label{examples}
+
+\subsection{Parallel Programming}
+\subsubsection{OpenCL}
 The development of the full OpenCL language using only the extension mechanisms described above provides evidence of the power of this approach.
@@
 
-\subsubsection{Parallel Programming}
-\paragraph{OpenCL}
-\subparagraph{Annotation and Extension Inference}
-In addition to type annotations, OpenCL normally asks for additional annotations in a number of other situations.
+%\subparagraph{Annotation and Extension Inference}
+%In addition to type annotations, OpenCL normally asks for additional annotations in a number of other situations.
 Users can annotate functions that meet certain requirements  with the \verb|__kernel| attribute, indicating that they are callable from the host.
@@
 An example of this can be seen in Listing \ref{mapout}, where the use of the \verb|double| type triggers the insertion of the \verb|cl_khr_f64| extension.
-
-\paragraph{Partitioned Global Address Spaces}
+%
+\subsubsection{Partitioned Global Address Spaces}
 A number of recent languages designed for clusters use a partitioned global address space model, including UPC \cite{upc}, Chapel \cite{chapel} and others.
@@
 
+\subsection{General-Purpose Abstractions}
 \paragraph{Object Models and Message Passing}
@@
 
-\paragraph{Functional Parallelism}
+\subsubsection{Functional Datatypes}
+
+\subsubsection{Functional Parallelism}
 Another promising approach for parallel programming is in automatic parallelization of purely functional programs.
@@
 
-\subsubsection{Other Use Cases in HPC}
-\paragraph{Interoperability Layers}
+\subsection{Interoperability Layers}
 The design criteria of \textbf{continuity} and \textbf{interoperability} require consideration of the large body of codes written in a variety of existing languages, across a number of paradigms.
@@
 
-\paragraph{Specialized Optimizations}
-In many cases, specialized code optimizations requires statically tracking invariants throughout a program.
+%\paragraph{Specialized Optimizations}
+%In many cases, specialized code optimizations requires statically tracking invariants throughout a program.
 Often these optimizations can be encoded as a type system for this reason.
@@
 
-\paragraph{Domain-Specific Type Systems}
+\subsection{Domain-Specific Type Systems}
+\subsubsection{Units of Measure}
 Although not strictly related to HPC, a number of domain-specific type systems related to computational science can be implemented within Ace.

### 2013-11-15 ace-pldi14/ace-pldi14.tex
pldi14 sectionized sec 4 roughly

@@
 \lstinputlisting{listing9.py}
-\caption{\texttt{[ace.OpenCL]} A portion of the implementation of OpenCL pointer types implementing subscripting logic using the Ace extension mechanism, AT\&T.}
+\caption{\texttt{[]} A portion of the implementation of OpenCL pointer types implementing subscripting logic using the Ace extension mechanism, AT\&T.}
 \label{pointers}

### 2013-11-15 ace-pldi14/listing9.py

@@ -1,25 +1,36 @@
 import ace
 
-class PtrType(ace.Type):
-  def __init__(self, target_type, addr_space):
-    self.target_type = target_type
+class Ptr(ace.ActiveType):
+  def __init__(self, addr_space, target_t):
     self.addr_space = addr_space
+    self.target_t = target_type
         
-  def resolve_Subscript(self, context, node):
-    slice_type = context.resolve(node.slice)
-    if isinstance(slice_type, IntegerType):
-      return self.target_type
+  def type_Subscript(self, context, node):
+    slice_t = context.type(node.slice)
+    if isinstance(slice_t, Integer):
+      return self.target_t
     else:
-      raise TypeError('<error message>', node)
+      raise ace.TypeError('<error message>', node)
        
   def translate_Subscript(self, context, node):
-    value = context.translate(node.value)
-    slice = context.translate(node.slice)
-    return ace.copy_node(node, 
-      value=value, slice=slice,    
-      code=value.code + '[' + slice.code + ']')      
-  
-  # ...
-  
-def gptr(target_type):
-    return PtrType(target_type, "__global")
+    value_x = context.translate(node.value)
+    slice_x = context.translate(node.slice)
+    return context.target.Subscript(value_x, slice_x)
+
+  def type_BinOp(self, context, node):
+    if node.op == ast.Add:
+      right_t = context.type(node.right)
+      if isinstance(right_t, Integer):
+        return self
+    elif node.op == ast.Sub:
+      right_t = context.type(node.right)
+      if (isinstance(right_t, Ptr) and 
+          right_t.addr_space == self.addr_space):
+        return ptrdiff_t
+    else: raise ace.TypeError('<error message>', node)
+
+  def translate_BinOp(self, context, node):
+    left_x = context.translate(node.left)
+    right_x = context.translate(node.right)
+    return context.target.BinOp(left_x, 
+      node.operator, right_x)

### 2013-11-15 ace-pldi14/ace-pldi14.tex
pldi14 merged heathers changes

@@
         }
+
+ % Heather-added packages for the fancy table
+ \usepackage{longtable}
+ \usepackage{booktabs}
+ \usepackage{pdflscape}
+ \usepackage{colortbl}%
+ \newcommand{\myrowcolour}{\rowcolor[gray]{0.925}}
+ \usepackage{wasysym}
 
@@
 Decorators in Python are simply syntactic sugar for applying the decorator function directly to the function  being decorated \cite{python}.
-In other words, line 3 could be replaced by inserting the following  statement on line 9\todo{figure out how to center this}:
-
-\verb|map = ace.fn(clx.base, clx.opencl)(map)|
+In other words, line 3 could be replaced by inserting the following  statement on line 9:
+\vspace{-0.2cm}
+\begin{verbbox}
+map = ace.fn(clx.base, clx.opencl)(map)
+\end{verbbox}
+\begin{figure}[h!]
+\centering
+\theverbbox
+\end{figure}
+\vspace{-0.2cm}
 
@@
 \section{Related Work}\label{related}
+\parindent0pt
+%
 \begin{figure*}
-\begin{tabular}[t]{ l | l | c c c c c }
-\textbf{Approach} & \textbf{Examples} & Library & {Extensible} & Extensible & Composable & Alternative \\
-\hline
-Active Types & Ace & x & ~ & x & x & x\\
-Desugaring & SugarJ, Sugar* & ~ & x & ~ & ~ & ~\\
-Rule Injection & Qi, Typed Racket/Clojure, A? & 1+2 & ~ & x & ~ & ~\\
-Static metaprogramming & Scala macros, OJ, OC++ & ~ & ~ & ~ & x & ~\\
-Cross-compilation & Delite & x & ~ & ~ & ~ & x\\
-EDSL frameworks & ? & x & x & x & ~ & ~\\
-Type-specific literals & Wyvern & ~ & x & ~ & x & ~\\
-\end{tabular}
+\onecolumn
+\begin{longtable}{l l@{}l c@{}c c@{}c c@{}c c@{}c c@{}c}
+%\toprule%
+
+&&  &&  && {\bfseries Extensible} && {\bfseries Extensible} &&  && {\bfseries Alternative}\\
+
+{\bfseries Approach} && {\bfseries Examples} && {\bfseries Library} && {\bfseries Syntax} && {\bfseries Type System} && {\bfseries Compositional} && {\bfseries Targets}  \\
+
+\cmidrule(l){1-1} \cmidrule(l){2-3} \cmidrule(l){4-5} \cmidrule(l){6-7} \cmidrule(l){8-9} \cmidrule(l){10-11} \cmidrule(l){12-13}
+
+\endhead
+
+Active Types && Ace && \CIRCLE && \Circle && \CIRCLE && \CIRCLE && \CIRCLE \\
+
+\myrowcolour%
+Desugaring && SugarJ \cite{erdweg2011sugarj}, Sugar* && \Circle && \CIRCLE && \Circle && \Circle && \Circle \\
+
+Rule Injection && Qi \cite{qi}, Typed Racket \cite{TypedScheme2008}, A? && 1+2 && \Circle && \CIRCLE && \Circle && \Circle \\
+
+\myrowcolour%
+Static macros /  && Scala macros \cite{ScalaMacros2013}, MorphJ \cite{MorphJ2011}, && \Circle && \Circle && \Circle && \CIRCLE && \Circle \\
+
+\myrowcolour
+Metaprogramming &&  OJ \cite{OpenJava2000}, Template Haskell \cite{template-haskell} && && && && && \\
+
+Cross-Compilation && Delite \cite{Delite2011} && \CIRCLE && \Circle && \Circle && \Circle && \CIRCLE \\
+
+\myrowcolour%
+EDSL Frameworks && ? && \CIRCLE && \CIRCLE && \CIRCLE && \Circle && \Circle \\
+
+Type-Specific Literals && Wyvern \cite{globaldsl13} && \Circle && \CIRCLE && \Circle && \CIRCLE && \Circle \\
+\end{longtable}
 \caption{Comparison to related approaches}
+\twocolumn
 \end{figure*}
@@
 
- Future work: integrate with something like scalad http://lampwww.epfl.ch/~hmiller/scala2013/resources/pdfs/paper8.pdf.
+ Future work: integrate with something like scalad http://lampwww.epfl.ch/~hmiller/scala2013/resources/pdfs/paper8.pdf. \todo{fix python lang reference citation}
 
@@
 
-\bibliographystyle{abbrvnat}
+\bibliographystyle{abbrv}
 

### 2013-11-15 ace-pldi14/ace-pldi14.tex
pldi14 active types, typechecking sections

@@
 The visitor pattern typically used in object-oriented languages implements essentially the same protocol.
-This does not work for an extensible language: new cases need to be added, and the corresponding logic must be included.
-In Ace, new types reuse existing forms, so this requires that the logic inside each case be modifiable by extensions (in a safely composable way).
+This does not work for an extensible language because new cases cannot be added to the types and functions (in a modular, safely composable manner).
 
@@
 \verb|e1[e2]|), Ace defers control over typechecking and translation to the active type of a designated subexpression.
-The choice of subexpression is fixed by the language, motivated by considerations of Python's semantics for the same form.
-Collectively, these choices comprise the \emph{dispatch protocol} of Ace.
+The choice of subexpression is determined by Ace's fixed \emph{dispatch protocol}, which is designed taking into consideration Python's semantics for the same form.
 Below are examples of these choices.
-Due to space constraints, we do not show the full protocol, consisting of a rule for each compound form in the grammar \cite{pythonast}.
+Due to space constraints, we do not show the full protocol.
 \begin{itemize}
@@
 \item Responsibility over {\bf simple assignment statements}, is handed to the type of the variable on the left (which, as we will see below, is determined by the active base).
-If this type does not provide special assignment semantics, the base handles it.
-%Destructuring assignment is also supported by a somewhat more complex protocol.\todo{clarify}
+If this type does not provide special assignment semantics, the base must handle it.%Destructuring assignment is also supported by a somewhat more complex protocol.\todo{clarify}
 \end{itemize}
 
-\subsubsection{Active Typechecking}
+\subsubsection{Typechecking}
+When typechecking a compound expression or statement, the Ace compiler calls a method of the form \verb|type_|$X$ on the active type that has responsibility over it according to the dispatch protocol, where $X$ is typically the name of the syntactic form, taken from the Python grammar \cite{pythonast}.
+For example, the typing logic for the \verb|Subscript| and \verb|BinOp| forms when operating on OpenCL pointers are shown beginning on lines \ref{pointers}.8 and \ref{pointers}.20, respectively.
+The typing methods are provided a \verb|context| and the abstract syntax tree of the \verb|node| being considered.
+The context is an instance of \verb|ace.Context| and tracks information like variable types, as coordinated by the base.
+It also includes a method \verb|type| that can be used to assign  type to subexpressions as needed.
+For example, line \ref{pointers}.8 checks if the type of the slice is an integer type.
+\verb|Integer| is simply another class in the \verb|clx| library that implements the \verb|ace.ActiveType| interface.
+Specific integer types like \verb|int| and \verb|ptrdiff_t| are instances of this class (details not shown).
+Note that the subexpressions are not \emph{automatically} assigned types.
+This is useful because it allows types to contextually reinterpret them to support more complex constructs (e.g.
+pattern matching on datatypes, see Sec.
+\ref{datatypes}).
+
+If a type error is detected, \verb|ace.TypeError| can be raised with an appropriate error message and a reference to the node that produced it (so that Ace can extract line and column numbers).
+We note that error messages are an important and sometimes overlooked facet of {ease-of-use} \cite{marceau2011measuring}.
+A common frustration with general-purpose abstraction mechanisms is that they can produce  verbose and cryptic error messages revealing implementation details of the abstraction being encoded (e.g.
+template metaprogramming \cite{template-metaprogramming}).
+
 \begin{codelisting}
 \lstinputlisting{listing9.py}
-\caption{\texttt{[]} A portion of the implementation of OpenCL pointer types implementing subscripting logic using the Ace extension mechanism, AT\&T.}
+\caption{\texttt{[}in \texttt{examples/clx.py]} The active type family \texttt{Ptr} implements the semantics of OpenCL pointer types.}
 \label{pointers}
 \end{codelisting}
-When the compiler encounters an expression, it must first verify its validity by either assigning it a type or raising a meaningful type error.
-Rather than defining fixed logic for this, Ace defers control to the {type} of the \emph{primary operand} according to a {\em dispatch protocol}.
-Because types are metalanguage instances of user-defined types derived from \verb|ace.Type| (cf.
-Section 2.5), dispatch corresponds to calling a method of this base class.
-
-Let us again consider the data-parallel map example of Section 2.
-When \verb|map| is compiled on Line 9 of Listing \ref{compscript}, the first two argument types are \verb|gptr(double)|.
-This type is an instance of the user-defined \verb|OpenCL.PtrType| which inherits from \verb|ace.Type|.
-When the compiler encounters the expression \verb|input[gid]| on Line 6, the dispatch protocol is to defer control to the type of \verb|input| by invoking the method named \verb|resolve_Subscript|.
-
-The relevant portion of \verb|OpenCL.PtrType| as well as \verb|gptr| is shown in Listing \ref{pointers}.
-The \verb|resolve_Subscript| method on Line 8 receives a context and the syntax tree of the node being considered.
-The context contains information about variables in scope and other contextual information, and also exposes a method, \verb|resolve_type|, that this method uses to recursively resolve the types of other subexpressions, here the slice which will be \verb|gid|, as needed.
-The context will have that \verb|gid| is the machine-dependent integer type \verb|size_t| as discussed in Section 2.
-On Line 10, it confirms that this type is an instance of an integer type and thus, it assigns the whole expression, \verb|input[gid]|, the target type of the pointer, \verb|double|.
-Had the function attempted to index \verb|input| using a non-integer expression, the method would take the other branch of the conditional and raise a type error, with a custom error message, on Line 13.
-We note that error messages are an important component of \textbf{ease-of-use} \cite{marceau2011measuring}.
-Indeed, a widely-reported frustration with C++ is that it produces overly verbose and cryptic error messages, particularly when templates are used in clever ways.
-
-\subsubsection{Active Translation}
+\subsubsection{Translation}
 Once typechecking a method is complete, the compiler must subsequently translate each Ace source expression into an expression in the target language.
@@
 
-&&  &&  && {\bfseries Extensible} && {\bfseries Extensible} &&  && {\bfseries Alternative}\\
+&&  &&  && {\bfseries Extensible} && {\bfseries Extensible} && {\bfseries Extensions} && {\bfseries Alternative}\\
 
@@
 \myrowcolour%
-Desugaring && SugarJ \cite{erdweg2011sugarj}, Sugar* && \Circle && \CIRCLE && \Circle && \Circle && \Circle \\
+Desugaring && SugarJ \cite{erdweg2011sugarj}, Sugar* \cite{sugarstar} && \Circle && \CIRCLE && \Circle && \Circle && \Circle \\
 
@@
 \myrowcolour%
-Static macros /  && Scala macros \cite{ScalaMacros2013}, MorphJ \cite{MorphJ2011}, && \Circle && \Circle && \Circle && \CIRCLE && \Circle \\
+Static macros /  && Scala \cite{ScalaMacros2013}, MorphJ \cite{MorphJ2011}, OJ \cite{OpenJava2000} && \Circle && \Circle && \Circle && \CIRCLE && \Circle \\
 
 \myrowcolour
-Metaprogramming &&  OJ \cite{OpenJava2000}, Template Haskell \cite{template-haskell} && && && && && \\
+Metaprogramming &&  Template Haskell \cite{template-haskell} && && && && && \\
 

### 2013-11-15 ace-pldi14/listing9.py

@@ -1,5 +1,3 @@
-import ace
-
 class Ptr(ace.ActiveType):
   def __init__(self, addr_space, target_t):
     self.addr_space = addr_space
@@ -17,20 +15,23 @@ class Ptr(ace.ActiveType):
     slice_x = context.translate(node.slice)
     return context.target.Subscript(value_x, slice_x)
 
-  def type_BinOp(self, context, node):
-    if node.op == ast.Add:
+  def type_BinOp_left(self, context, node):
+    if isinstance(node.operator, ast.Add):
       right_t = context.type(node.right)
       if isinstance(right_t, Integer):
         return self
-    elif node.op == ast.Sub:
+    elif isinstance(node.operator, ast.Sub):
       right_t = context.type(node.right)
       if (isinstance(right_t, Ptr) and 
           right_t.addr_space == self.addr_space):
         return ptrdiff_t
-    else: raise ace.TypeError('<error message>', node)
+    raise ace.TypeError('<error message>', node)
 
-  def translate_BinOp(self, context, node):
+  def translate_BinOp_left(self, context, node):
     left_x = context.translate(node.left)
     right_x = context.translate(node.right)
     return context.target.BinOp(left_x, 
       node.operator, right_x)
+
+  # type_BinOp_right and translate_BinOp_right 
+  # (not shown) are symmetric

### 2013-11-15 ace-pldi14/Ptr.py
pldi14 progress

@@ -1,27 +1,3 @@
-class Cplx(ace.ActiveType):
-  def __init__(self, t):
-    if not isinstance(t, Numeric):
-      raise ace.InvalidType("<error message>")
-    self.t = t
-
-  def translate_type(self, context, target):
-    return target.Vec(2, self.t.translate_type(context))
-
-  def type_BinOp_left(self, context, node):
-    right_t = context.type(node.right)
-    if isinstance(right_t, Numeric):
-      return Cplx(c99_binop_t(self.t, right_t))
-    elif isinstance(right_t, Cplx):
-      return Cplx(c99_binop_t(self.t, right.t))
-
-  def translate_BinOp_left(self, context, target, node):
-    right_t = context.type(node.right)
-    target = context.target
-    if isinstance(right_t, Numeric):
-      left_x = context.translate(node.left)
-      right_x = context.translate(node.right)
-      return target.BinOp()
-
 class Ptr(ace.ActiveType):
   def __init__(self, addr_space, t):
     self.addr_space = addr_space

### 2013-11-16 ace-pldi14/ace-pldi14.tex
pldi14 progressssssslowwwwwwwww

@@
 In other words, line 3 could be replaced by inserting the following  statement on line 9:
-\vspace{-0.2cm}
+\vspace{-0.28cm}
 \begin{verbbox}
@@
 \end{figure}
-\vspace{-0.2cm}
+\vspace{-0.28cm}
 
@@
 
-\todo{move this to related work??}Type assignment to generic functions is similar in some ways to template specialization in C++.
-In effect, both the template header and type parameters at call sites are being generated automatically by Ace.
-This significantly simplifies a sophisticated feature of C++ and introduces it to targets like OpenCL and C, which do not support templates.
-Other uses for C++ templates (and the preprocessor) are subsumed by the metaprogramming features discussed above\todo{cite template metaprogramming paper}.
+Type assignment to generic functions is similar in some ways to template specialization in C++.
+In effect, both a template header and type parameters at call sites are being generated automatically by Ace.
+This simplifies a sophisticated feature of C++ and enables its use with other targets like OpenCL.
+%Other uses for C++ templates (and the preprocessor) are subsumed by the metaprogramming features discussed above\todo{cite template metaprogramming paper}.
 
@@
 For developers experienced with the usual OpenCL or CUDA workflow, the fact that this can be accomplished in a total of 6 statements may be surprising.
-This simplicity is enabled largely due to  implicit tracking of types throughout the code.
+This simplicity is possibly largely due to  implicit tracking of types throughout the code.
 First, \verb|numpy| keeps track of the type, shape and order of its arrays.
-The type of \verb|input|, for example, is \verb|numpy.double| by default, its shape is \verb|(1024,)| and its order is row-major by default.
+The type of \verb|input|, for example, is \verb|numpy.double| by default, its shape is \verb|(1024,1024)| and its order is row-major by default.
 The \verb|pyopencl| library that our mechanism is built atop uses this information to automatically call the underlying OpenCL host API function for transferring byte arrays to the device without requiring the user to calculate the size.
-Our wrapper of \verb|pyopencl| retains this information in the Python representations of the device arrays, \verb|d_input| and \verb|d_output|.
-An Ace type parameterized can use this metadata to typecheck and generate code for operations like indexing in Ace functions\todo{rephrase}.
+Our wrapper of \verb|pyopencl| further retains this information in the Python wrappers around the device arrays, \verb|d_input| and \verb|d_output|.
+The Ace active type of such wrappers can be designated to be an instance of \verb|clx.NPArray| parametermized by this metadata.
+This type knows how to typecheck and translate operations like indexing automatically (see Sec.
+\ref{atypes}).
 
@@
 Both the \verb|map| and \verb|negate| functions must be explicitly specialized with the appropriate types using string manipulation techniques.
-Higher order functions are still not available, and must be simulated by string manipulation.
-That approach also does not permit the use any of the language extensions that Ace enables (in this case, the only interesting use of the extension mechanism was by the Ace type for \verb|numpy| arrays, which ensures that indexing respects ordering; see Sec.
+Higher order functions are still not available, and must also be simulated by string manipulation.
+That approach also does not permit the use any of the language extensions that Ace enables (e.g.
+the type for \verb|numpy| arrays, which ensures that indexing respects ordering; see Sec.
 \ref{examples} for more interesting possibilities).
@@
 \section{Extensibility}\label{att}
-The core of Ace consists of about 1500 lines of Python code implementing its primary concepts: generic functions, concrete functions, active bases, active types and active targets.
+The core of Ace consists of about 1500 lines of Python code implementing its primary concepts: generic functions, concrete functions, active types, active bases and active targets.
 The latter three comprise Ace's extension mechanism.
 Extensions provide semantics to, and govern the compilation of, Ace functions, rather than logic in Ace's core.
-Indeed, the name ``Ace'' might itself be an acronym: it is an ``active compilation environment''.
+%Indeed, the name ``Ace'' might itself be an acronym: it is an ``active compilation environment''.
 
 \subsection{Active Types}\label{atypes}
-Active types are the primary means for extending Ace, and motivate active bases, so we will describe them first.
-An active type, as mentioned above, is an instance of a class implementing the \verb|ace.ActiveType| interface.
-Listing \ref{pointers} shows an example of such a class: the \verb|clx.Ptr| class used in Listing \ref{compscript}, which implements the logic of OpenCL's pointer types.
-The class takes two arguments in its constructor: an address space and the target type of the pointer.
-OpenCL pointers support several operations: they can be indexed by an integer type to extract the value being pointed to; one can perform pointer arithmetic on them; and two pointers can be subtracted to produce a value of type \verb|ptrdiff_t| (a device-dependent integer type) if they are in the same address space \cite{opencl11}.
+Active types are the primary means for extending Ace with new abstractions.
+An active type, as mentioned previously, is an instance of a class implementing the \verb|ace.ActiveType| interface.
+Listing \ref{cplx} shows an example of such a class: the \verb|clx.Cplx| class used in Listing \ref{compscript}, which implements the logic of complex numbers.
+The constructor takes as a parameter any numeric type in \verb|clx| (line \ref{cplx}.2).
 
+\subsubsection{Dispatch Protocol}
+\begin{codelisting}
+\lstinputlisting{listing9.py}
+\caption{\texttt{[}in \texttt{examples/clx.py]} The active type family \texttt{Ptr} implements the semantics of OpenCL pointer types.}
+\label{cplx}
+\end{codelisting}
 In a compiler for a monolithic language, there would be a \emph{syntax-directed} protocol governing typechecking and translation.
-For example, in a functional language, one would declare datatypes that captured all forms of types and expressions, and the typechecker would perform exhaustive case analysis over the expression forms, implementing the logic described above in the cases for subscript (\verb|e1[e2]|) and addition (\verb|e1 + e2|).
+In a compiler written in a functional language, for example, one would declare datatypes that captured all forms of types and expressions, and the typechecker would perform exhaustive case analysis over the expression forms, so all the semantics would be implemented in one place.
 The visitor pattern typically used in object-oriented languages implements essentially the same protocol.
-This does not work for an extensible language because new cases cannot be added to the types and functions (in a modular, safely composable manner).
+This does not work for an extensible language because new cases and logic need to be added modularly, in a safely composable manner.
 
 Instead of taking a syntax-directed approach, the Ace compiler's typechecking and translation phases take a \emph{type-directed approach}.
-When encountering a compound expression (e.g.
-\verb|e1[e2]|), Ace defers control over typechecking and translation to the active type of a designated subexpression.
-The choice of subexpression is determined by Ace's fixed \emph{dispatch protocol}, which is designed taking into consideration Python's semantics for the same form.
+When encountering a compound term (e.g.
+\verb|e1[e2]|), the compiler defers control over typechecking and translation to the active type of a designated subexpression (e.g.
+\verb|e1|) determined by Ace's fixed \emph{dispatch protocol}.
 Below are examples of these choices.
@@
 \ref{related} for a discussion.}
-\item Responsibility over {\bf simple assignment statements}, is handed to the type of the variable on the left (which, as we will see below, is determined by the active base).
+\item Responsibility over \textbf{constructor calls} (\verb|[t](e1, ..., en)|), where \verb|t| is a \emph{compile-time Python expression} evaluating to an active type, is handed to that type.
+If \verb|t| evaluates to a \emph{family} of types, like \verb|clx.Cplx|, the active type is first generated via a class method, as discussed below.
+%\item Responsibility over {\bf simple assignment statements}, is handed to the type of the variable on the left (which, as we will see below, is determined by the active base).
 If this type does not provide special assignment semantics, the base must handle it.%Destructuring assignment is also supported by a somewhat more complex protocol.\todo{clarify}
@@
 \subsubsection{Typechecking}
-When typechecking a compound expression or statement, the Ace compiler calls a method of the form \verb|type_|$X$ on the active type that has responsibility over it according to the dispatch protocol, where $X$ is typically the name of the syntactic form, taken from the Python grammar \cite{pythonast}.
-For example, the typing logic for the \verb|Subscript| and \verb|BinOp| forms when operating on OpenCL pointers are shown beginning on lines \ref{pointers}.8 and \ref{pointers}.20, respectively.
-The typing methods are provided a \verb|context| and the abstract syntax tree of the \verb|node| being considered.
-The context is an instance of \verb|ace.Context| and tracks information like variable types, as coordinated by the base.
-It also includes a method \verb|type| that can be used to assign  type to subexpressions as needed.
-For example, line \ref{pointers}.8 checks if the type of the slice is an integer type.
-\verb|Integer| is simply another class in the \verb|clx| library that implements the \verb|ace.ActiveType| interface.
-Specific integer types like \verb|int| and \verb|ptrdiff_t| are instances of this class (details not shown).
-Note that the subexpressions are not \emph{automatically} assigned types.
-This is useful because it allows types to contextually reinterpret them to support more complex constructs (e.g.
-pattern matching on datatypes, see Sec.
+When typechecking a compound expression or statement, the Ace compiler temporarily hands control to the object selected by the dispatch protocol by calling the method \verb|type_|$X$, where $X$ is  the name of the syntactic form, taken from the Python grammar \cite{pythonast} (appended with a suffix in some cases).
+
+For example, if \verb|c| is a complex number, then \verb|c.ni| and \verb|c.i| are its non-imaginary and imaginary components, respectively.
+These expressions are of the form \verb|Attribute|, so the typechecker calls \verb|type_Attribute| (line \ref{cplx}.7).
+This method receives the compilation context, \verb|context|, and the abstract syntax tree of the expression, \verb|node| and must return a type assignment for the node, or raise an \verb|ace.TypeError| if there is an error.
+In this case, a type assignment is possible if the attribute name is either \verb|"ni"| or \verb|"i"|, and an error is raised otherwise (lines \ref{cplx}.8-\ref{cplx}.10).
+We note that error messages are an important and sometimes overlooked facet of {ease-of-use} \cite{marceau2011measuring}.
+A common frustration with using general-purpose abstraction mechanisms to encode an abstraction is that they can produce  verbose and cryptic error messages that reflect the implementation details, rather than the semantics.
+%Active types permit direct encoding of semantics and customization of error messages.
+
+Complex numbers also support binary arithmetic operations partnered with both other complex numbers and with non-complex numbers, treating them as if their imaginary component is zero.
+The typechecking rules for this logic is implemented on lines \ref{cplx}.17-\ref{cplx}.29.
+Because arithmetic operations are meant to be symmetric, the dispatch protocol checks both subexpressions for support, favoring the left to ensure that the semantics remain deterministic.
+In either position, the implementation begins by recursively assigning a type to the other operand in the current context via the \verb|context.type| method (line \ref{cplx}.24).
+If supported, it applies the C99 rules for arithmetic operations to determine the resulting type (\cite{c99}, not shown).
+
+Finally, a complex number can be constructed inside an Ace function using Ace's special constructor form: \verb|[clx.Cplx](3,4)| represents $3+4i$, for example.
+The term within the braces is evaluated at \emph{compile-time}.
+Because \verb|clx.Cplx| evaluates not to an active type, but to a class, this form is assigned a type by handing control to the class object via the \emph{class method} \verb|type_New|.
+It operates as expected, extracting the types of the two arguments to construct an appropriate complex number type (lines \ref{cplx}.50-\ref{cplx}.57), raising a type error if the arguments cannot be promoted to a common type according to the rules of C99 or if two arguments have not been provided (an exercise for the reader: modify this method to also allow a single argument for when the imaginary part is 0).
+
+%Note that in each example, subexpressions are not \emph{automatically} assigned types.
+Types must be requested from the context.
+This is useful because it allows types to reinterpret subexpressions of terms they control, to supporting more complex constructs (e.g.
+pattern matching on functional datatypes, see Sec.
 \ref{datatypes}).
 
-If a type error is detected, \verb|ace.TypeError| can be raised with an appropriate error message and a reference to the node that produced it (so that Ace can extract line and column numbers).
-We note that error messages are an important and sometimes overlooked facet of {ease-of-use} \cite{marceau2011measuring}.
-A common frustration with general-purpose abstraction mechanisms is that they can produce  verbose and cryptic error messages revealing implementation details of the abstraction being encoded (e.g.
-template metaprogramming \cite{template-metaprogramming}).
+\subsubsection{Translation}
+Once typechecking a method is complete, the compiler enters the translation phase, where terms in the target language are generated from Ace terms.
+Terms in the target language are generated by calling methods of the \emph{active targets}.
+The translation phase operates similarly to typechecking, using the dispatch protocol to invoke methods named \verb|trans_|$X$.
+These methods have access to the context and node just as during typechecking, as well as the target.
 
-\begin{codelisting}
-\lstinputlisting{listing9.py}
-\caption{\texttt{[}in \texttt{examples/clx.py]} The active type family \texttt{Ptr} implements the semantics of OpenCL pointer types.}
-\label{pointers}
-\end{codelisting}
-\subsubsection{Translation}
-Once typechecking a method is complete, the compiler must subsequently translate each Ace source expression into an expression in the target language.
-This has been OpenCL in the examples thus far, but we describe a generalization of this in the next section.
+Translation of terms of the form \verb|c.ni| and \verb|c.i| operates via \verb|trans_Attribute|.
+Because we are implementing complex numbers as vector types in OpenCL, we generate an access to the left element, accessed via the attribute \verb|s0|, to access the non-imaginary component and and \verb|s1| to access the imaginary component.
+The target provides a programmatic API for generating code in the target language via its methods, as can be seen.
+Translation can assume typechecking succeeded, so the implementation assumes that if the attribute is not \verb|ni|, it is \verb|i|.
 
@@
 The context also provides some support for non-local insertions, such as new top-level type declarations, imports and helper code (not shown).
+
+\subsection{Active Targets}\label{atargets}
+
+\subsubsection{Target Types}
+Every active type must translate to a unique  \emph{target type} -- a type in the target language (dynamically typed target languages can be thought of as only having a single type, \verb|dyn| \cite{pfpl}).
+Target types are constructed via methods of a provided {active target}, called \verb|target| throughout this listing.
+Here, a complex number translates to a pair of values, each having the type that is the target type of \verb|t|, the numeric type that was passed into the constructor (lines \ref{cplx}.7-\ref{cplx}.8).
+The \verb|tuple| method returns a representation of an actual type (not an active type) in the target language as the return value of the \verb|tuple| method (e.g.
+2-tuples of integers compile to the vector type \verb|int2| in OpenCL).
+This design still allows the complex number type being defined here to supports any target for which tuples can be defined directly however (e.g.
+CUDA, Python, etc.).
+
+\todo{nathan's work on representational consistency?}
 
@@
 Indeed, to even use \verb|double| in OpenCL, an appropriate \verb|#pragma| must be inserted. The base detects uses of \verb|double| anywhere inside the function
-
-\subsection{Active Targets}\label{atargets}
-
-If no active target is explicitly provided, as in our example, the active base  provides a default target.
-The default target provided by \verb|clx.base| is \verb|clx.opencl|, but other targets, like \verb|clx.cuda|, a vendor-specific language similar to OpenCL, as well as Python itself (\verb|ace.python|), can also be defined in libraries.
-One base can support multiple targets.
-We will discuss targets in Sec.
-\ref{targets}.
-
-Thus far, we have been discussing the \verb|OpenCL| module in our examples.
-This module faithfully implements all the types and operations of the OpenCL kernel language, a portable standard for writing low-level code on multi-core processors and accelerators \cite{opencl11}.
-The functions we wrote have used these primitives with the \verb|OpenCL.OpenCL| backend, so the translation has been direct and no run-time overhead has been introduced.
-Thus, as described so far, Ace has affirmatively resolved the question it was originally conceived to address, discussed in Section 2.1.
 

### 2013-11-16 ace-pldi14/listing8.py

@@ -2,7 +2,7 @@ import listing1, listing2, examples.clx as clx, numpy
 
 clx.opencl.ctx = clx.Context.for_device(0, 0)
 
-input = numpy.ones((1024,))
+input = numpy.ones((1024,1024))
 d_input = clx.to_device(input)
 d_output = clx.alloc(like=input)
 

### 2013-11-16 ace-pldi14/listing9.py

@@ -1,64 +1,63 @@
 class Cplx(ace.ActiveType):
   def __init__(self, t):
     if not isinstance(t, Numeric):
-      raise ace.InvalidType("<error message>")
+      raise ace.InvalidTypeError("<error message>")
     self.t = t
 
-  def translate_type(self, context, target):
-    return target.Vec(2, self.t.translate_type(context))
-
-  def type_BinOp_left(self, context, node):
-    right_t = context.type(node.right)
-    if isinstance(right_t, Numeric):
-      return Cplx(c99_binop_t(self.t, right_t))
-    elif isinstance(right_t, Cplx):
-      return Cplx(c99_binop_t(self.t, right.t))
-
-  def translate_BinOp_left(self, context, target, node):
-    right_t = context.type(node.right)
-    target = context.target
-    if isinstance(right_t, Numeric):
-      left_x = context.translate(node.left)
-      right_x = context.translate(node.right)
-      return target.BinOp()
-
-class Ptr(ace.ActiveType):
-  def __init__(self, addr_space, t):
-    self.addr_space = addr_space
-    self.t = t
-
-  def translate_type(self, context):
-    t_x = self.t.translate_type(context)
-    return target.Ptr(self.addr_space, t_x)
-        
-  def type_Subscript(self, context, node):
-    slice_t = context.type(node.slice)
-    if isinstance(slice_t, Integer):
+  def type_Attribute(self, context, node):
+    if node.attr == 'ni' or node.attr == 'i':
       return self.t
-    else:
-      raise ace.TypeError('<error message>', node)
-       
-  def translate_Subscript(self, context, node):
+    raise ace.TypeError("<error message>", node)
+
+  def trans_Attribute(self, context, target, node):
     value_x = context.translate(node.value)
-    slice_x = context.translate(node.slice)
-    return context.target.Subscript(value_x, slice_x)
+    a = 's0' if node.attr == 'ni' else 's1'
+    return target.Attribute(value_x, a)
 
   def type_BinOp_left(self, context, node):
-    if isinstance(node.operator, ast.Add):
-      right_t = context.type(node.right)
-      if isinstance(right_t, Integer):
-        return self
-    elif isinstance(node.operator, ast.Sub):
-      right_t = context.type(node.right)
-      if self == right_t:
-        return ptrdiff_t
-    raise ace.TypeError('<error message>', node)
[…]

### 2013-11-16 ace-pldi14/ace-pldi14.tex
pldi14 sec 3 draft1 finally done HOLY GOD NO TIME LEFT

@@
 
-Translation of terms of the form \verb|c.ni| and \verb|c.i| operates via \verb|trans_Attribute|.
-Because we are implementing complex numbers as vector types in OpenCL, we generate an access to the left element, accessed via the attribute \verb|s0|, to access the non-imaginary component and and \verb|s1| to access the imaginary component.
-The target provides a programmatic API for generating code in the target language via its methods, as can be seen.
-Translation can assume typechecking succeeded, so the implementation assumes that if the attribute is not \verb|ni|, it is \verb|i|.
+As seen in Listing \ref{mapout}, we are implementing complex numbers internally using OpenCL vector types, like \verb|int2|.
+Let us look first at \verb|trans_New| on lines \ref{cplx}.54-\ref{cplx}.60, where new complex numbers are translated to vector literals by invoking \verb|VecLit| from the active target, which internally will generate the necessary OpenCL string for that form.
+The \verb|trans_type| method of the \verb|ace.ActiveType| is used to associated a type in the target language with an active type.
+For it to be possible to reason compositionally about the correctness of compilation, all complex numbers must translate to terms in the target language that have this target type \cite{tdc}.
+Ace supports a mode where this \emph{representational consistency} is dynamically checked during compilation (this requires that the active target know how to assign types to terms in the target language, which can be done only for our OpenCL target as of this writing).
 
-It does so by again applying the dispatch protocol to call a method of the form \verb|translate_|$X$, where $X$ is the syntactic form of the expression.
-This method is responsible for returning a copy of the expression's ast node with an additional attribute, \verb|code|, containing the source code of the translation, represented here as a string though it may also be represented in a structured manner.
-In our example, it is simply a direct translation to the corresponding OpenCL attribute access (Line 20), using the recursively-determined translations of the operands (Lines 16-17).
-More sophisticated abstractions may insert arbitrarily complex statements and expressions during this phase.
-The context also provides some support for non-local insertions, such as new top-level type declarations, imports and helper code (not shown).
-
-\subsection{Active Targets}\label{atargets}
-
-\subsubsection{Target Types}
-Every active type must translate to a unique  \emph{target type} -- a type in the target language (dynamically typed target languages can be thought of as only having a single type, \verb|dyn| \cite{pfpl}).
-Target types are constructed via methods of a provided {active target}, called \verb|target| throughout this listing.
-Here, a complex number translates to a pair of values, each having the type that is the target type of \verb|t|, the numeric type that was passed into the constructor (lines \ref{cplx}.7-\ref{cplx}.8).
-The \verb|tuple| method returns a representation of an actual type (not an active type) in the target language as the return value of the \verb|tuple| method (e.g.
-2-tuples of integers compile to the vector type \verb|int2| in OpenCL).
-This design still allows the complex number type being defined here to supports any target for which tuples can be defined directly however (e.g.
-CUDA, Python, etc.).
-
-\todo{nathan's work on representational consistency?}
+The translation methods for attributes and binary operations proceed in a straightforward manner.
+The context provides a method, \verb|trans|, for recursively determining the translation of subexpressions as needed.
+Of note is that the translation methods can assume that typechecking succeeded.
+For example, the implementation of \verb|trans_Attribute| assumes that if \verb|node.attr| is not \verb|'ni'| then it must have been \verb|'i'| on line \ref{cplx}.14, consistent with the implementation of \verb|type_Attribute| above it.
+Typechecking and translation are separated into two methods to emphasize that typechecking is not target-dependent and to allow for typing rules that require generating types for hypothetical terms, where the overhead of the translation might be better avoided.
 
@@
 Each generic function is associated with an active base, which is an object implementing the \verb|ace.ActiveBase| interface.
-The active base specifies the \emph{base semantics} of that function -- the semantics of statements and expressions that do not have a clear ``primary'' subexpression.
+The active base specifies the \emph{base semantics} of that function.
+It controls the semantics of statements and expressions that do not have a clear ``primary'' subexpression for the dispatch protocol to defer to.
+A base is handed control over typechecking of statements and expressions in the same way as active types: via \verb|type_|$X$ and \verb|trans_|$X$ methods.
+It also initializes the context and governs the semantics of variables.
+Each function can have a distinct base semantics.
 
-For example, the type and translation of a numeric literal is determined by the active base.
-The active base \verb|clx.base|, for example, assigns integer literals the type \verb|clx.int32| while floating point literals have type \verb|clx.double|, to be consistent with the semantics of OpenCL and C99.
-The \verb|clx.base| object is an instance of \verb|clx.Base|.
-This class is written so that this choice can be made differently if desired.
-For example, the base \verb|clx.Base(flit_t=clx.float)| changes the base semantics so that floating point literals have type \verb|clx.float|.
-This is useful because some OpenCL devices do not support double-precision floating point, or impose a significant performance penalty for its use.
-For applications where this higher precision is not necessary, this is convenient.
-Similarly, a base where integer literals are represented using true integers, rather than 32-bit integers, might be useful in  applications where avoiding accidental overflows is more important than performance.
-True integers can be implemented as an active type (not shown).
+Literals are the most prominent form given their semantics by an active base.
+Our example active base, \verb|clx.base|, assigns integer literals the type \verb|clx.int32| while floating point literals have type \verb|clx.double|, consistent with the semantics of OpenCL and C99.
+The \verb|clx.base| object is an instance of \verb|clx.Base| that is pre-constructed for convenience.
+Alternative bases can be generated via different constructor arguments.
+For example, \verb|clx.Base(flit_t=clx.float)| is a base semantics where floating point literals have type \verb|clx.float|.
+This is useful because some OpenCL devices do not support double-precision numbers, or impose a significant performance penalty for their use.
+Indeed, to even use the \verb|double| type in OpenCL, a flag must be toggled by a \verb|#pragma| (The OpenCL target we have implemented inserts this automatically when needed, along with some other flags, and annotations like \verb|kernel|).
+Similarly, for some applications, avoiding accidental overflows is more important than performance.
+Arbitrary-precision integers can be implemented as an active type (not shown) and the base generates the constructor call form, described above, around integers to support this.
 
-Indeed, to even use \verb|double| in OpenCL, an appropriate \verb|#pragma| must be inserted. The base detects uses of \verb|double| anywhere inside the function
+Another role that the active base can play is to provide operators that do not need to be imported.
+The semantics of \verb|get_global_id| and \verb|printf|, for example, are provided by \verb|clx.base|.
+In fact, the base provides a richer semantics for them than the underlying implementation, as we will show in the following section.
+It is not strictly necessary that the base provide these -- they could have been imported from \verb|clx| and used by their qualified name.
+A base which does not provide them without import can be generated by passing the \verb|primitives=False| flag in.
 
-\subsubsection{Composability and Interoperability}\label{safety}
-Because a type can only exert control over typechecking and translation of operations where an expression of that type is the primary operand, extensions defined using AT\&T can not interfere with one another by construction.
-This does not imply that there are no \textbf{interoperability} issues to consider, however.
-A type may need to know about other types (e.g.
-pointer types need to know about integer types) and if this is done without consideration of future extensions, it may be difficult to integrate data produced by one type system with another.
-These issues cannot easily be addressed by a language design, however.
+\subsection{Active Targets}\label{atargets}
+An active target, which we show being used in the translation logic described above, is an instance of \verb|ace.ActiveTarget|.
+It has two primary roles in Ace: 1) it provides an API for code generation to a target language, providing term and type constructors for use during translation (and, to support representational consistency checks, a type system implementation); 2) it provides a bridge to the underlying interoperability API when a generic or concrete function is launched directly from Python, as described in Sec.
+\ref{implicit}.
 
-\section{Examples}\label{examples}
+An active type or active base can support multiple targets if desired by simply examining it during translation (e.g.
+by performing capability checks using Python's \verb|hasattr| function) and providing alternative code paths for different combinations of capabilities.
+Alternatively, an active backend can implement the interfaces available in a different back end and generate code for them in a different way.
+Our implementation of OpenCL's type system as a collection of active types and an active base does not explicitly support cross-compilation to a CUDA target (though it would be possible to modify it to do so), but our CUDA target does provide partial support for the OpenCL code generation API.
+
+\subsection{Compositional Reasoning}\label{safety}
+Every statement and expression in an Ace function is governed by exactly one active type, determined by the dispatch protocol, or by the single active base associated with the Ace function.
+The representational consistency check ensures that compilation is successful without requiring that types that abstract over other types (e.g.
+container types) know about their internal implementation details.
+Together, this permits compositional reasoning about the semantics of Ace functions even in the presence of many extensions.
+This stands on contrast to many prior approaches to extensibility, where extensions could insert themselves into the semantics in conflicting ways, making the order of imports matter (see Sec.
+\ref{related}).
+
+\section{Expressiveness}\label{examples}
+Ace comes with a target, base and type implementing Python itself: \verb|ace.python.python|, \verb|ace.python.base| and \verb|ace.python.dyn|.
+These can be supplemented by additional active types and used as the foundation for writing actively-typed Python functions.
+These functions can either be compiled ahead-of-time to an untyped Python file for execution, or be immediately executed with just-in-time compilation, just like the OpenCL examples we have shown.
+Some of the examples in the next section support this target in addition to the OpenCL target that we have been focusing on in this paper.
+
+\subsection{Growing a Statically-Typed Python Inside an Ace}
+Ace comes with a target, base and type implementing Python itself: \verb|ace.python.python|, \verb|ace.python.base| and \verb|ace.python.dyn|.
+These can be supplemented by additional active types and used as the foundation for writing actively-typed Python functions.
+These functions can either be compiled ahead-of-time to an untyped Python file for execution, or be immediately executed with just-in-time compilation, just like the OpenCL examples we have shown.
+Some of the examples in the next section support this target in addition to the OpenCL target that we have been focusing on in this paper.
 
@@
 \section{Discussion}\label{discussion}
+Python does not truly prevent extensions from interfering with one another because it lacks, e.g., data hiding mechanisms.
+One type could change the implementation of another by replacing it's methods, for example.
+
 Static type systems are powerful tools for programming language design and implementation.

### 2013-11-16 ace-pldi14/listing9.py

@@ -10,7 +10,7 @@ class Cplx(ace.ActiveType):
     raise ace.TypeError("<error message>", node)
 
   def trans_Attribute(self, context, target, node):
-    value_x = context.translate(node.value)
+    value_x = context.trans(node.value)
     a = 's0' if node.attr == 'ni' else 's1'
     return target.Attribute(value_x, a)
 
@@ -30,8 +30,8 @@ class Cplx(ace.ActiveType):
 
   def trans_BinOp(self, context, target, node):
     r_t = context.type(node.right)
-    l_x = context.translate(node.left)
-    r_x = context.translate(node.right)
+    l_x = context.trans(node.left)
+    r_x = context.trans(node.right)
     make = lambda a, b: target.VecLit(
       self.trans_type(self, target), a, b)
     binop = lambda a, b: target.BinOp(
@@ -39,7 +39,7 @@ class Cplx(ace.ActiveType):
     si = lambda a, i: target.Attribute(a, 's'+str(i))
     if isinstance(r_t, Numeric):
       return make(binop(si(l_x, 0), r_x), si(r_x, 1))
-    elif isinstance(right_t, Cplx):
+    elif isinstance(r_t, Cplx):
       return make(binop(si(l_x, 0), si(r_x, 0)),
         binop(si(l_x, 1), si(r_x, 1)))
 

### 2013-11-16 ace-pldi14/ace-pldi14.tex
pldi14 sec 4 intro

@@
 \section{Expressiveness}\label{examples}
-Ace comes with a target, base and type implementing Python itself: \verb|ace.python.python|, \verb|ace.python.base| and \verb|ace.python.dyn|.
-These can be supplemented by additional active types and used as the foundation for writing actively-typed Python functions.
-These functions can either be compiled ahead-of-time to an untyped Python file for execution, or be immediately executed with just-in-time compilation, just like the OpenCL examples we have shown.
-Some of the examples in the next section support this target in addition to the OpenCL target that we have been focusing on in this paper.
+Thus far, we have focused mainly on the OpenCL target and shown examples of fairly low-level active types: those that implement OpenCL's primitives (e.g.
+\verb|clx.Ptr|), extend them in simple but convenient ways (e.g.
+\verb|clx.Cplx|) and those that make interactive execution across language boundaries safe and convenient (e.g.
+\verb|clx.NPArray|).
+Ace was first conceived to answer the question: \emph{can we build a statically-typed language with the semantics of a C but the syntax and ease-of-use of a Python?} We submit that these examples (along with a large-scale simulator framework built using them, which we do not have space to detail here) have answered this question in the affirmative.
+
+But Ace has proven useful for more than low-level tasks like programming a GPU with OpenCL.
+We now describe several interesting extensions that implement semantics of primitives drawn from a range of different language paradigms, to justify our claim that these mechanisms are highly expressive.
 
@@
 These functions can either be compiled ahead-of-time to an untyped Python file for execution, or be immediately executed with just-in-time compilation, just like the OpenCL examples we have shown.
-Some of the examples in the next section support this target in addition to the OpenCL target that we have been focusing on in this paper.
+Some of the examples in the next section support this target in addition to the OpenCL target we have focused on thus far.
 

### 2013-11-16 ace-pldi14/ace-pldi14.tex
Added FP example code

@@
 But Ace has proven useful for more than low-level tasks like programming a GPU with OpenCL.
-We now describe several interesting extensions that implement semantics of primitives drawn from a range of different language paradigms, to justify our claim that these mechanisms are highly expressive.
+We now describe several interesting extensions that implement the semantics of primitives drawn from a range of different language paradigms, to justify our claim that these mechanisms are highly expressive.
 
@@
 These functions can either be compiled ahead-of-time to an untyped Python file for execution, or be immediately executed with just-in-time compilation, just like the OpenCL examples we have shown.
-Some of the examples in the next section support this target in addition to the OpenCL target we have focused on thus far.
+Many of the examples in the next section support this target in addition to the OpenCL target we have focused on thus far.
+
+\subsection{Functional Datatypes}
+\begin{codelisting}
+\lstinputlisting{datatypes_t.py}
+\caption{x.}
+\label{datatypest}
+\end{codelisting}
+\begin{codelisting}
+\lstinputlisting{datatypes.py}
+\caption{y.}
+\label{datatypes}
+\end{codelisting}
+HAIOADJASDASD
 

### 2013-11-16 ace-pldi14/datatypes.py

@@ -1,0 +1,28 @@
+class DT(object):
+	pass
+
+class DT_Empty(DT):
+	def __init__(self, data):
+		pass
+
+class DT_Leaf(DT):
+	def __init__(self, data):
+		self.data = data
+
+class DT_Node(DT):
+	def __init__(self, data):
+		self.data = data
+
+def depth_gt_2(x):
+	if isinstance(x, DT_Node):
+		if isinstance(x.data[0], DT_Node):
+			return True
+		if isinstance(x.data[1], DT_Node):
+			return True
+	return False
+
+if __name__ == "__main__":
+	my_lil_tree = DT_Node(DT_Empty(), DT_Empty())
+	my_big_tree = DT_Node(my_lil_tree, my_lil_tree)
+	assert not depth_gt_2(my_lil_tree)
+	assert depth_gt_2(my_big_tree)

### 2013-11-16 ace-pldi14/datatypes_t.py

@@ -1,0 +1,28 @@
+import ace, ace.python as py, examples.fp as fp
+
+def Tree(a):
+	t = fp.Datatype()
+	t.cases = {
+		'empty': fp.unit,
+		'leaf': a,
+		'node': fp.tuple(t, t)
+	}
+	t.close()
+	return t
+
+DT = Tree(py.dyn)
+
+@ace.fn(py.Base(trailing_return=True))[[DT]]
+def depth_gt_2(t):
+	t.case({
+		dt.node(dt.node(_), _): True,
+		dt.node(_, dt.node(_)): True,
+		_: False
+	})
+
+@ace.fn(py.Base(main=True))[[]]
+def __main__():
+	my_lil_tree = dt.node(dt.empty, dt.empty)
+	my_big_tree = dt.node(my_lil_tree, my_lil_tree)
+	assert not depth_gt_2(my_lil_tree)
+	assert depth_gt_2(my_big_tree)

### 2013-11-16 ace-pldi14/ace-pldi14.tex
Finished examples

@@
 \lstinputlisting{datatypes_t.py}
-\caption{x.}
+\caption{\texttt{[datatypes\_t.py]} An example using statically-typed functional datatypes.}
 \label{datatypest}
@@
 \lstinputlisting{datatypes.py}
-\caption{y.}
+\caption{\texttt{[datatypes.py]} The dynamically-typed Python code generated by running \texttt{acec datatypes\_t.py}.}
 \label{datatypes}
 \end{codelisting}
-HAIOADJASDASD
+Listing \ref{datatypest} shows an example of the use of statically-typed functional datatypes alongside the Python implementation just described.
+It shows two syntactic conveniences that were not mentioned previously: 1) if no target is provided to \verb|ace.fn|, the base can provide a default target (here, \verb|ace.python.python|; 2) A concrete function can be generated immediately by providing a type assignment to \verb|ace.fn|.
+Listing \ref{datatypest} generates Listing \ref{datatypes}.
 
-\subsection{Parallel Programming}
-\subsubsection{OpenCL}
-The development of the full OpenCL language using only the extension mechanisms described above provides evidence of the power of this approach.
-However, to be truly useful, the mechanism must be able to express a wide array of higher-level primitive abstractions.
-We briefly describe a number of other abstractions that are possible using this mechanism.
-Many of these are currently available in existing languages either via libraries or as primitives of some specialized language.
-A study comparing a language-based concurrency solution for Java with an equivalent, though less clean, library-based solution found that language support is preferable but leads to many of the issues we have described \cite{cave2010comparing}.
+Lines \ref{datatypest}.3-\ref{datatypest}.11 define a function that generates a recursive algebraic datatype representing a tree given a name and another Ace type for the data at the leaves.
+This type implemented by the active type family \verb|fp.Datatype|.
+Case names and their types are provided programmatically on lines \ref{datatypest}.5-\ref{datatypest}.10.
+These lines also show two more active type families: units (the type containing just one value), and immutable pairs.
+Line \ref{datatypest}.13 calls this function with the Ace type for dynamic Python values, \verb|py.dyn|, to generate a type, aliased \verb|DT|.
+This type is implemented using class inheritance when the target is Python, as seen on lines \ref{datatypes}.1-\ref{datatypes}.13.
+For C-like targets, a \verb|union| type can be used (not shown).
+
+The generic function \verb|depth_gt_2| demonstrates two features.
+First, the base has been setup to treat the final expression in a top-level branch as the return value of the function, consistent with typical functional languages.
+Second, the \verb|case| ``method'' on line \ref{datatypest}.17 creatively reuses the syntax for dictionary literals to express a nested pattern matching construct.
+The patterns to the left of the colons are not treated as expressions (that is, a type and translation is never recursively assigned to them).
+The active type implements the standard algorithm for ensuring that the cases cover all possibilities and are not redundant \cite{pfpl}.
+If the final case were omitted, for example, the algorithm would statically indicate an error, just as in statically-typed functional languages like ML and Haskell.
+
+\subsection{Nested Data Parallelism}
+Functional constructs have shown promise as the basis for a nested data-parallel parallel programming model.
+Many powerful parallel algorithms can be specified by composing primitives like \verb|map| and \verb|reduce| on persistent data structures like lists, trees and records.
+Data dependencies are directly encoded in these specifications, and automatic code generation techniques have shown promise in using this information to automatically execute these specifications on concurrent hardware and networks.
+The Copperhead programming language, for example, is based in Python as well and allows users to express functional programs over lists using common functional primitives, compiling them to CUDA code \cite{catanzaro2011copperhead}.
+By combining active types and the metaprogramming facilities described in Sec.
+2 to implement optimizations, like fusion, on the typed syntax trees available from concrete functions, these same techniques can be implemented as an Ace library as well.
+
+\subsection{Structs, Unions, Records and Objects}
+Data structures like structs, unions, records and objects can all be implemented as Ace type families  parameterized by a dictionary mapping field names to types.
+Each family differs slightly in the semantics of the operations available.
+Structs support mutation, for example, while records do not.
+Object systems typically introduce additional logic for calling methods, binding special variables like \verb|this|, and accessing information through an inheritance hierarchy.
+Listing \ref{ooclxfppy} shows an example of each of these.
+The prototypic object system delegates to a backing record if a field is not available in the struct \cite{self}.
+This example also demonstrates cross-compilation to C99, supported for a subset of operations in \verb|clx|.
+All three abstractions are implemented as simple \verb|struct|s, as can be seen in Listing \ref{ooclxfpc}.
+This example also demonstrates the use of argument annotations to generate concrete functions (line \ref{ooclxfppy}.9), which is a feature only available when running Ace in Python 3.
+
+\subsection{Distributed Programming}
+Many distributed programming abstractions can be understood as implementing object models with complex forms of dynamic dispatch.
+For example, in Charm++, dynamic dispatch involves message passing over a dynamically load-balanced network \cite{kale1993charm++}.
+While Charm++ is a sophisticated system, and we do not anticipate that implementing it as an Ace extension would be easy, it is possible to do so by targeting a system like Adaptive MPI, which exposes the Charm++ runtime system \cite{kale2009charm++}.
+Erlang and other message-passing and actor-based systems can also be considered in this manner -- processes can be thought of as objects, channels as methods and messages as calls.
 
@@
 \subsubsection{Partitioned Global Address Spaces}
-A number of recent languages designed for clusters use a partitioned global address space model, including UPC \cite{upc}, Chapel \cite{chapel} and others.
+
+A number of recent languages designed for distributed programming on clusters use a partitioned global address space model, including UPC \cite{upc}, Chapel \cite{chapel} and others.
 These languages provide first-class support for accessing data transparently across a massively parallel cluster.
@@
 
-\subsection{General-Purpose Abstractions}
-\paragraph{Object Models and Message Passing}
-Classes in object-oriented systems can be understood as inducing a type parameterized by a static type specification that emits code to perform dynamic dispatch by some protocol.
-In Ace, the object system would be a subclass of \verb|ace.Type| (e.g.
-\verb|JavaObject|), the specification would be a compile-time Python object (e.g.
-a dictionary mapping names to types, as is done with OpenCL \verb|struct|s, augmented with information about subtyping and methods in more complex systems).
-The code for checking field and method accesses against the specification would be be in \verb|resolve_Attribute| method and the dynamic dispatch mechanism would be emitted in \verb|translate_Attribute|.
-That is, Ace supports arbitrary object models, rather than fixing a single one that may not fit all needs, as in almost all other object-oriented languages.
-
-Many parallel programming abstractions can be understood as implementing object models with complex forms of dynamic dispatch.
-For example, in Charm++, dynamic dispatch involves message passing over a dynamically load-balanced network \cite{kale1993charm++}.
-While Charm++ is a very sophisticated system, and we do not anticipate that implementing it as an Ace extension would be easy, it is entirely possible to do so using a system like Adaptive MPI that exposes its runtime system \cite{kale2009charm++}.
-Erlang and other message-passing based systems can also be considered in this manner -- processes can be thought of as objects, channels as methods and messages as calls.
-
-\subsubsection{Functional Datatypes}
-
-\subsubsection{Functional Parallelism}
-Another promising approach for parallel programming is in automatic parallelization of purely functional programs.
-These languages express computations that do not explicitly manipulate memory.
-They rely instead on the composition of primitives like \verb|map| and \verb|reduce| that transform components of persistent data structures like lists, trees and maps.
-Data dependencies are thus directly encoded in programs, and automatic parallelization techniques have shown promise in using this information to automatically schedule these parallel programs on concurrent hardware and networks.
-
-The Copperhead language, for example, is based on Python as well and allows users to express functional programs over lists using common functional primitives, compiling them to CUDA code \cite{catanzaro2011copperhead}.
-We believe that Ace can express precisely the same programming model, implemented as an extension.
-We have prototyped support for a number of common persistent data structures, including algebraic datatypes, in Ace.
-Perhaps surprisingly, Python's imperative syntax does not preclude writing programs in a reasonable functional style (due largely to its support for tuples).
-
-\subsection{Interoperability Layers}
-The design criteria of \textbf{continuity} and \textbf{interoperability} require consideration of the large body of codes written in a variety of existing languages, across a number of paradigms.
-Although it is often possible to call from one language to another using a foreign function interface (FFI), this is almost never natural or statically safe.
-In Ace, however, extensions could be written that internalize the foreign language's type system (note that dynamically-typed languages can be considered to have a single type, often called $\mathtt{dyn}$) and emit code that hides the FFI from users, achieving \textbf{verifiability} and \textbf{ease-of-use}.
-
-%\paragraph{Specialized Optimizations}
-%In many cases, specialized code optimizations requires statically tracking invariants throughout a program.
-Often these optimizations can be encoded as a type system for this reason.
-For instance, a course project using Ace  implemented substantial portions of the GPU-specific optimizations described in \cite{yang2010gpgpu} as a library, using types to track affine transformations of the global ID in order to construct a summary of the memory access patterns of the kernel.
-This information can be used both for single-kernel optimization (as in \cite{yang2010gpgpu}) and for future research on cross-kernel fusion and other optimizations.
-
-\subsection{Domain-Specific Type Systems}
 \subsubsection{Units of Measure}
-Although not strictly related to HPC, a number of domain-specific type systems related to computational science can be implemented within Ace.
+A number of domain-specific type systems can be implemented within Ace as well.
 For example, prior work has considered tracking units of measure (e.g.
 grams) statically to validate scientific code \cite{conf/cefp/Kennedy09}.
-This cannot easily be implemented using existing abstraction mechanisms because this information should only be maintained statically to avoid excessive run-time overhead associated with tagging.
-The Ace extension mechanism allows this information to be tracked in the type system, but not included during translation.
-
-\paragraph{Instrumentation}
-Several sophisticated feedback-directed optimizations and adaptive run-time protocols require instrumenting code in various ways.
-The extension mechanism combined with support for patching classes dynamically using Python enables granular instrumentation that can consider both the syntactic form of an operation as well as its constituent types, easing the implementation of such tools.
-
-This ability could also be used to collect data useful for more rigorous usability and usage studies of languages and abstractions, and we plan on following up on this line of research going forward.
+This cannot easily be implemented using many existing abstraction mechanisms because this information should only be maintained statically to avoid excessive run-time overhead associated with tagging/boxing, and the typechecking logic is reasonably complex.
+The Ace extension mechanism allows this information to be tracked in the type system, but not included during translation, by a mechanism similar to how address space information is tracked with pointers.
+Because Ace extensions can use Python, the needed logic can be implemented directly.
 
 \section{Related Work}\label{related}
+\begin{codelisting}[t]
+\lstinputlisting{ooclxfp.py}
+\caption{\texttt{[ooclxfp.py]} An example combining structs and immutable records using a prototype-based object system, cross-compiled to C99. Uses Python 3 argument annotations.}
+\label{ooclxfppy}
+\end{codelisting}
+\begin{codelisting}[t]
+\lstinputlisting[style=OpenCL]{ooclxfp.c}
+\caption{\texttt{[ooclxfp.c]} The C99 code generated by running \texttt{acec ooclxfp.py}.}
+\label{ooclxfpc}
+\end{codelisting}
+
 \parindent0pt

### 2013-11-16 ace-pldi14/datatypes.py

@@ -1,28 +1,29 @@
 class DT(object):
-	pass
+  pass
 
 class DT_Empty(DT):
-	def __init__(self, data):
-		pass
-
+  pass
+  
 class DT_Leaf(DT):
-	def __init__(self, data):
-		self.data = data
+  def __init__(self, data):
+    self.data = data
 
 class DT_Node(DT):
-	def __init__(self, data):
-		self.data = data
+  def __init__(self, data):
+    self.data = data
 
 def depth_gt_2(x):
-	if isinstance(x, DT_Node):
-		if isinstance(x.data[0], DT_Node):
-			return True
-		if isinstance(x.data[1], DT_Node):
-			return True
-	return False
+  if isinstance(x, DT_Node):
+    if isinstance(x.data[0], DT_Node):
+      r = True
+    elif isinstance(x.data[1], DT_Node):
+      r = True
+  elif True:
+    r = False
+  return r
 
 if __name__ == "__main__":
-	my_lil_tree = DT_Node(DT_Empty(), DT_Empty())
-	my_big_tree = DT_Node(my_lil_tree, my_lil_tree)
-	assert not depth_gt_2(my_lil_tree)
-	assert depth_gt_2(my_big_tree)
+  my_lil_tree = DT_Node(DT_Empty(), DT_Empty())
+  my_big_tree = DT_Node(my_lil_tree, my_lil_tree)
+  assert not depth_gt_2(my_lil_tree)
+  assert depth_gt_2(my_big_tree)

### 2013-11-16 ace-pldi14/datatypes_t.py

@@ -1,28 +1,28 @@
 import ace, ace.python as py, examples.fp as fp
 
-def Tree(a):
-	t = fp.Datatype()
-	t.cases = {
-		'empty': fp.unit,
-		'leaf': a,
-		'node': fp.tuple(t, t)
-	}
-	t.close()
-	return t
+def Tree(name, a):
+  t = fp.Datatype(name)
+  t.cases = {
+    'Empty': fp.unit,
+    'Leaf': a,
+    'Node': fp.tuple(t, t)
+  }
+  t.close()
+  return t
 
-DT = Tree(py.dyn)
+DT = Tree('DT', py.dyn)
 
 @ace.fn(py.Base(trailing_return=True))[[DT]]
 def depth_gt_2(t):
-	t.case({
-		dt.node(dt.node(_), _): True,
-		dt.node(_, dt.node(_)): True,
-		_: False
-	})
+  t.case({
+    DT.Node(DT.Node(_), _): True,
+    DT.Node(_, DT.Node(_)): True,
+    _: False
+  })
 
 @ace.fn(py.Base(main=True))[[]]
 def __main__():
-	my_lil_tree = dt.node(dt.empty, dt.empty)
-	my_big_tree = dt.node(my_lil_tree, my_lil_tree)
-	assert not depth_gt_2(my_lil_tree)
-	assert depth_gt_2(my_big_tree)
+  my_lil_tree = DT.Node(DT.Empty, DT.Empty)
+  my_big_tree = DT.Node(my_lil_tree, my_lil_tree)
+  assert not depth_gt_2(my_lil_tree)
+  assert depth_gt_2(my_big_tree)

### 2013-11-16 ace-pldi14/ooclxfp.c

@@ -1,0 +1,16 @@
+typedef struct {
+	int x;
+} A;
+
+typedef struct {
+	float y;
+} B;
+
+typedef struct {
+	A self;
+	B proto;
+} C;
+
+float test(C input) {
+	return C.self.x + C.proto.y;
+}

### 2013-11-16 ace-pldi14/ace-pldi14.tex
pldi14 final submission

@@
 First, an extension mechanism itself may require modifying a base language with constructs for defining, importing and using extensions.
-The extension mechanism is not itself a library, so it faces many of the same extrinsic issues as other new languages like HJ, leading to a ``bootstrapping'' problem.
-The extension mechanisms available today also have several intrinsic problems related to safety and expressiveness that require technical solutions before it would be appropriate to widely rely on them.
-We evaluate related work in Section \ref{related}.
+The extension mechanism is not itself a library, so it faces a ``bootstrapping'' problem: it must overcome many of the same extrinsic issues as other new languages like HJ.
+The extension mechanisms available today also have some intrinsic problems related to safety and expressiveness that require technical solutions before it would be appropriate to widely rely on them.
 
@@
 \ref{examples} by showing a diverse collection of abstractions drawn from different language paradigms that can be implemented as orthogonal libraries in Ace.
-We include functional datatypes, objects, macros, and typesafe format strings and regular expressions.
+We include functional datatypes, objects, and several other abstractions.
 In Sec.
@@
 A common workflow for \emph{professional end-user programmers} (e.g.
-scientists and engineers \cite{peus}) is to use a simple scripting language for orchestration, small-scale data analysis and visualization and call into a low-level language for performance-critical sections.
+scientists and engineers) is to use a simple scripting language for orchestration, small-scale data analysis and visualization and call into a low-level language for performance-critical sections.
 Python is both designed for this style of use and widely adopted for such tasks \cite{sanner1999python,nguyen2010survey}.
@@
 \begin{itemize}
-\item Responsibility over {\bf attribute access} (\texttt{e.attr}), {\bf subscripting} (\texttt{e[e1]}\todo{is it n-ary?}) and \textbf{calls} (\verb|e(e1, ..., en)|) and {\bf unary operations} (e.g.
+\item Responsibility over {\bf attribute access} (\texttt{e.attr}), {\bf subscripting} (\texttt{e[e1]}) and \textbf{calls} (\verb|e(e1, ..., en)|) and {\bf unary operations} (e.g.
 \verb|-e|) is handed to the type recursively assigned to \texttt{e}.
@@
 \subsubsection{Typechecking}
-When typechecking a compound expression or statement, the Ace compiler temporarily hands control to the object selected by the dispatch protocol by calling the method \verb|type_|$X$, where $X$ is  the name of the syntactic form, taken from the Python grammar \cite{pythonast} (appended with a suffix in some cases).
+When typechecking a compound expression or statement, the Ace compiler temporarily hands control to the object selected by the dispatch protocol by calling the method \verb|type_|$X$, where $X$ is  the name of the syntactic form, taken from the Python grammar \cite{python} (appended with a suffix in some cases).
 
@@
 In either position, the implementation begins by recursively assigning a type to the other operand in the current context via the \verb|context.type| method (line \ref{cplx}.24).
-If supported, it applies the C99 rules for arithmetic operations to determine the resulting type (\cite{c99}, not shown).
+If supported, it applies the C99 rules for arithmetic operations to determine the resulting type (not shown).
 
@@
 The \verb|trans_type| method of the \verb|ace.ActiveType| is used to associated a type in the target language with an active type.
-For it to be possible to reason compositionally about the correctness of compilation, all complex numbers must translate to terms in the target language that have this target type \cite{tdc}.
+For it to be possible to reason compositionally about the correctness of compilation, all complex numbers must translate to terms in the target language that have this target type.
 Ace supports a mode where this \emph{representational consistency} is dynamically checked during compilation (this requires that the active target know how to assign types to terms in the target language, which can be done only for our OpenCL target as of this writing).
@@
 The patterns to the left of the colons are not treated as expressions (that is, a type and translation is never recursively assigned to them).
-The active type implements the standard algorithm for ensuring that the cases cover all possibilities and are not redundant \cite{pfpl}.
+The active type implements the standard algorithm for ensuring that the cases cover all possibilities and are not redundant.
 If the final case were omitted, for example, the algorithm would statically indicate an error, just as in statically-typed functional languages like ML and Haskell.
@@
 Listing \ref{ooclxfppy} shows an example of each of these.
-The prototypic object system delegates to a backing record if a field is not available in the struct \cite{self}.
+The prototypic object system delegates to a backing record if a field is not available in the struct.
 This example also demonstrates cross-compilation to C99, supported for a subset of operations in \verb|clx|.
@@
 While Charm++ is a sophisticated system, and we do not anticipate that implementing it as an Ace extension would be easy, it is possible to do so by targeting a system like Adaptive MPI, which exposes the Charm++ runtime system \cite{kale2009charm++}.
-Erlang and other message-passing and actor-based systems can also be considered in this manner -- processes can be thought of as objects, channels as methods and messages as calls.
+%Erlang and other message-passing and actor-based systems can also be considered in this manner -- processes can be thought of as objects, channels as methods and messages as calls.
 
@@
 %
-\subsubsection{Partitioned Global Address Spaces}
 
@@
 
-\parindent0pt
+%\parindent0pt
 %
@@
 \myrowcolour%
-Desugaring && SugarJ \cite{erdweg2011sugarj}, Sugar* \cite{sugarstar} && \Circle && \CIRCLE && \Circle && \Circle && \Circle \\
+Desugaring && SugarJ \cite{erdweg2011sugarj} && \Circle && \CIRCLE && \Circle && \Circle && \Circle \\
 
-Rule Injection && Qi \cite{qi}, Typed Racket \cite{TypedScheme2008}, A? && 1+2 && \Circle && \CIRCLE && \Circle && \Circle \\
+Rule Injection && Qi, Typed Racket \cite{TypedScheme2008}, Xroma \cite{activelibraries} && some && \Circle && \CIRCLE && \Circle && \Circle \\
 
@@
 \myrowcolour
-Metaprogramming &&  Template Haskell \cite{template-haskell} && && && && && \\
+Metaprogramming &&  MetaML \cite{Sheard:1999:UMS}, Template Haskell && && && && && \\
 
@@
 \myrowcolour%
-EDSL Frameworks && ? && \CIRCLE && \CIRCLE && \CIRCLE && \Circle && \Circle \\
+%EDSL Frameworks && ? && \CIRCLE && \CIRCLE && \CIRCLE && \Circle && \Circle \\
 
-Type-Specific Literals && Wyvern \cite{globaldsl13} && \Circle && \CIRCLE && \Circle && \CIRCLE && \Circle \\
+%Type-Specific Literals && Wyvern \cite{globaldsl13} && \Circle && \CIRCLE && \Circle && \CIRCLE && \Circle \\
 \end{longtable}
-\caption{Comparison to related approaches}
+\caption{Comparison to related approaches to language-internal extensibility.}
 \twocolumn
 \end{figure*}
-\subsection{Active Libraries}
+%\subsection{Active Libraries}
+
 Libraries that contain compile-time logic have been called {\it active libraries} in prior proposals \cite{activelibraries}.
 A number of  projects, such as Blitz++, have taken advantage of the C++ preprocessor and template-based metaprogramming system to implement domain-specific optimizations \cite{veldhuizen2000blitz++}.
-In Ace, we replace these brittle mini-languages with a general-purpose language and significantly expand the notion of active libraries by consideration of types as objects in the metalanguage.
-We thus call these types \emph{active types}.
+In Ace, we replace these brittle mini-languages with a general-purpose language, Python, and significantly expand the notion of active libraries by consideration of types, base semantics and target languages as values (in this case, objects) in the compile-time metalanguage.
 
-\subsection{Structural Polymorphism}
-Generic functions represent a novel strategy for achieving {\it function polymorphism} -- the ability to define functions that operate over more than a single type.
+Generic functions are a novel strategy for {\it function polymorphism} -- defining functions that operate over more than a single type.
 In Ace, generic functions are implicitly polymorphic and can be called with arguments of {\it any type that supports the operations used by the function}.
-This approach is related to structural polymorphism, however \cite{malayeri2009structural}.
-Structural types make explicit the requirements on a function, unlike generic functions.
+This approach is related to structural polymorphism \cite{malayeri2009structural}.
+Structural types make explicit the requirements on a function, unlike generic functions, which are only given singleton types because the requirements can be defined arbitrarily by extensions.
+Structural typing can be compared to the more \emph{ad hoc} approach taken by dynamically-typed languages like Python itself, sometimes called ``duck typing''.
+It is also comparable to the C++ template system, as discussed previously.
+Debugging type errors can be more difficult when there are many nested generic functions that were implicitly given type assignments.
+Type debuggers like \verb|scalad| could help make this easier \cite{scalad}.
 
-Structural typing can be compared to the more \emph{ad hoc} approach taken by dynamically-typed languages, sometimes called ``duck typing''.
-It is more flexible than the parametric polymorphism found in many functional languages and in languages like Java (which only allow polymorphic functions that are valid for {\it all} possible types), but is comparable to the C++ template system, as discussed previously.
-
-\subsection{Run-Time Indirection}
 {\it Operator overloading} \cite{vanWijngaarden:Mailloux:Peck:Koster:Sintzoff:Lindsey:Meertens:Fisker:acta:1975} and {\it metaobject dispatch} \cite{Kiczales91} are run-time protocols that translate operator invocations into function calls.
 The function is typically selected according to the type or value of one or more operands.
-These protocols share the notion of {\it inversion of control} with type-level specification.
-However, type-level specification is a {\it compile-time} protocol focused on enabling specialized verification and implementation strategies, rather than simply enabling run-time indirection.
+These protocols share the notion of {\it inversion of control} with our strategy.
+However, our strategy is a {\it compile-time} protocol where the typechecking and translation semantics are implemented for different operators.
+Our dispatch protocol was designed to be conceptually similar to Python's operator overloading protocol \cite{python}.
 
-\subsection{Term Rewriting Systems}
-Many languages and tools allow developers to rewrite expressions according to custom rules.
-These can broadly be classified as {\it term rewriting systems}.
-Macro systems, such as those characteristic of the LISP family of languages \cite{mccarthy1978history}, are the most prominent example.
-Some compile-time metaprogramming systems also allow users to manipulate syntax trees (e.g.
-MetaML \cite{Sheard:1999:UMS}), and external rewrite systems also exist for many languages.
-These differ in their direct exposure to syntax trees and their difficulties with propagating type information, since it is not directly encoded in the syntax.
-The AT\&T mechanism is a type-based mechanism that avoids these issues.
+There are several other language-internal extension mechanisms that have been described in the literature.
+These differ from our mechanism based primarily on active types in one of the following ways, summarized in the table in Figure 1:
+\begin{itemize}
+\item Our mechanism is itself implemented as a library, rather than as a language extension.
+\item Our mechanism reuses an existing language's syntax, rather than offering syntax extension support.
+As we saw in, for example, Listing \ref{datatypest}, this can still be reasonably natural.
+However, some mechanisms provide true syntax extensions.
+\item Our mechanism permits true type system extensions.
+New types are not merely aliases, nor must their rules be admissible in some base type system.
+Term rewriting and macro systems do not handle types at all.
+\item As described in Sec.
+\ref{safety}, our mechanism permits compositional reasoning.
+Techniques that allow global extensions to the syntax or semantics (e.g.
+SugarJ or rule injection systems like Xroma or Typed Racket) allow extensions so much control that they can interfere with one another or make reasoning about code difficult.
+\item Our mechanism allows users to control the target language of compilation from within libraries. Many other mechanisms are based fundamentally on rewriting mechanisms.
+\end{itemize}
 
-\subsection{Language Frameworks and \\Extensible Compilers}
 When the mechanisms available in an existing language prove insufficient, researchers and domain experts often design a new language.
 A number of tools have been developed to assist with this task, including compiler generators, language workbenches and domain-specific language frameworks (cf \cite{fowler2010domain}).
-Extensible compilers can be considered a form of language framework as well due to portability issues that using compiler extensions can introduce.
-It is difficult or impossible for these language-external approaches to achieve interoperability, as discussed above.
-
-\subsection{Extensible Languages}
-Extensible languages like SugarJ \cite{erdweg2011sugarj} afford some of the extensibility benefits of Ace, but are not targeted toward HPC.
-They have largely focused on syntactic extensibility, while Ace relies on a fixed syntax and emphasizes semantic extensibility.
-They also generally allow users to extend languages globally, which leads to conflicts when multiple extensions are used.
-AT\&T does not admit such conflicts.
+Extensible compilers can be considered a form of language framework as well due to the interoperability issues that relying on a compiler-specific extensions can introduce.
+It is difficult or impossible for these language-external approaches to achieve interoperability, because languages are not aware of each other's semantics.
 
 \section{Discussion}\label{discussion}
-Python does not truly prevent extensions from interfering with one another because it lacks, e.g., data hiding mechanisms.
-One type could change the implementation of another by replacing it's methods, for example.
-
 Static type systems are powerful tools for programming language design and implementation.
@@
 This simplifies and increases the performance of the run-time system, as errors need not be detected dynamically using tag checks and other kinds of assertions.
-Many parallel programming abstractions are defined in terms of, or benefit from, a type system that enforces a communication protocol, ensures the consistency of data and simplifies the dynamics of the run-time system (see Section \ref{usecases} for examples).
-Because \textbf{verifiability} and \textbf{performance} are key criteria and static typing is a core technique, Ace is fundamentally statically-typed.
 
-It is therefore a concern that most programming languages are {\em monolithic} -- a collection of primitives are given first-class treatment by the language implementation, and users can only creatively combine them to implement algorithms and abstractions of their design.
-Although highly-expressive general-purpose mechanisms have been developed (such as object systems or algebraic datatypes), these may not suffice when researchers or domain experts wish to evolve aspects of the type system, exert control over the representation of data, introduce specialized run-time mechanisms, or if defining an abstraction in terms of existing mechanisms is unnatural or verbose (in summary, to push the boundaries of \textbf{verifiability}, \textbf{performance} and \textbf{ease-of-use}).
-In these situations, it would be desirable to have the ability to modularly extend existing systems (\textbf{continuity}) with new compile-time logic (\textbf{extensibility}) and be assured that such extensions will never interfere with one another when used in the same program (\textbf{interoperability}).
-
-OpenCL is not necessarily the best tool for every job in high-performance computing.
-Indeed, HPC is an area where designing a set of primitives that satisfy all users has been particularly challenging, and it appears unlikely that a broad consensus will emerge given the variety of architectures, applications, scales and user communities that it serves, and the number of seemingly promising abstractions that emerge continuously targeting various subsets of this problem space.
-
-It is legitimate to ask, however, why dynamically-typed languages are so widely-used in HPC.
+It is legitimate to ask, however, why dynamically-typed languages are so widely-used in multiple domains.
 Although slow and difficult to reason about, these languages generally excel at satisfying the criteria of \textbf{ease-of-use}.
@@
 redundancy and the availability of reasonable defaults \cite{cordy1992hints}.
-Statically-typed languages, particularly those that HPC programmers are exposed to, are verbose, requiring explicit and often redundant type annotations on each function and variable declaration, separate header files, explicit template headers and  instantiation and other sorts of annotations.
-The dynamically-typed languages used in HPC, on the other hand, avoid most of this overhead by relying on support from the run-time system.
-Ace was first conceived to explore the question: \emph{does conciseness require run-time mechanisms, or can one develop a statically-typed language with the same low-level memory and execution model of C but syntactic overhead comparable to a high-level scripting language?
-}
+Statically-typed languages, particularly those that professional end-users are exposed to, are often verbose, requiring explicit and often redundant type annotations on each function and variable declaration, separate header files, explicit template headers and  instantiation and other sorts of annotations.
+Dynamically-typed languages, on the other hand, avoid most of this overhead by relying on support from the run-time system.
+Ace was first conceived to explore the question: \emph{does conciseness require run-time mechanisms, or can one develop a statically-typed language with the same usability profile?}
 
-Rather than designing a new syntax, or modifying the syntax of C, we chose to utilize, \emph{without modification}, the syntax of an existing language, Python.
-This choice was not arbitrary, but rather a key means by which Ace achieves both \textbf{ease-of-use} and \textbf{continuity}.
-Python's whitespace-delimited syntax is widely regarded as both concise and readable, and Python is amongst the most widely-adopted languages in computational science \cite{oliphant2007python}.
-By directly adopting Python's syntax, Ace's syntax is immediately \emph{familiar} and \emph{acceptable} to a significant segment of the intended audience.
-Moreover, a key benefit of adopting it without modifications is that any {tools} that handle Python source code, including parsers, editors, style checkers and documentation generators, can be used on Ace code without modification.
+Rather than designing a new syntax, or modifying the syntax of an existing language, we chose to utilize, \emph{without modification}, the syntax of an existing language, Python.
+This choice was not arbitrary, but rather a key means by which Ace satisfies extrinsic design criteria related to familiarity and tool support.
 Researchers often dismiss the importance of syntax.
-By using a well-developed syntax, they no longer need to worry about the equally ``trivial'' task of implementing tools for it.
+By repurposing a well-developed syntax, they no longer need to worry about the ``trivial'' task of implementing it.
 
-Professional end-users demand much from new languages and abstractions.
-In this paper, we began by generating a concrete set of design and adoption criteria that we hope will be of interest and utility to the research community.
-Based on these constraints, we designed a new language, Ace, making several pragmatic design decisions and introducing several novel techniques, including type propagation via generic functions, extensible type inference, active typechecking and translation and type-aware Python-Ace-OpenCL bindings to uniquely satisfy many of the criteria we discussed, particularly the three criteria that are typically overlooked in other languages.
-We validated the extension mechanism with a mature implementation of  the entirety of the OpenCL type system, as well as outlined a number of other use cases.
-Finally, we demonstrated that this language was useful in practice, drastically improving performance without negatively impacting the high-level scientific workflow of a large-scale neurobiological circuit simulation project.
+Most programming languages are {\em monolithic} -- a collection of primitives are given first-class treatment by the language implementation, and users can only creatively combine them to implement  abstractions of their design.
+Although highly-expressive general-purpose mechanisms have been developed (such as object systems or functional datatypes), these may not suffice when researchers or domain experts wish to evolve aspects of the type system, exert control over the representation of data, introduce specialized run-time mechanisms, if defining an abstraction in terms of existing mechanisms is unnatural or verbose, or if custom error messages are useful.
+In these situations, it would be desirable to have the ability to modularly extend existing systems with new compile-time logic and be assured that such extensions will never interfere with one another when used in the same program.
+
+Python does not truly prevent extensions from interfering with one another because it lacks, e.g., data hiding mechanisms.
+One type could change the implementation of another by replacing its methods, for example.
+But these kinds of conflicts do not occur ``innocently'', as conflicts between two libraries in SugarJ that use similar syntax might.
 
 Ace has some limitations at the moment.
-Debugging is only supported on the generated code, so if code generation introduces significant complexity, this can be an issue.
+Debuggers and other tools that rely not just on Python's syntax but also its semantics cannot be used directly, so if code generation introduces significant complexity that leads to bugs, this can be an issue.
 The OpenCL library we have implemented is a reasonably straightforward internalization of OpenCL itself, however, so debugging has not been a problem thusfar.
-We believe that active types can be useful to control debugging, and plan to explore this in the future.
-We will also further explore the use cases and case studies that we have described to validate the design we propose here.
-We hope that Ace will be developed further by the community to strengthen the foundations upon which new abstractions are implemented and deployed into the HPC professional end-user community.
+We believe that active types can be used to control debugging, and plan to explore this in the future.
+We have also not yet evaluated the feasibility of implementing more advanced type systems (e.g.
+linear type systems, dependent type systems, flow-dependent type systems) using Ace.
 
-We suggest three mutually-related {design criteria} that, unlike those in bold above, many languages and language-integrated abstractions have failed to adequately consider: \textbf{continuity}, \textbf{extensibility} and \textbf{interoperability}.
-These criteria encompass the intuitions that new abstractions will not be adopted in a vacuum, that programming systems must  support change, and that interacting components of an application or workflow should be able to make use of different abstractions naturally and without the possibility of conflict arising at their interface boundaries.
+Not all extensions will be useful.
+Indeed, some language designers worry that offering too much flexibility to users leads to abuse (this is, for example, widely credited as the reason why Java doesn't support operator overloading).
+We do not argue with this point.
+Instead, we argue that the potential for abuse must be balanced with the possibilities made available by a vibrant ecosystem of competing statically-typed abstractions that can be developed and deployed as libraries, and thus more easily evaluated in the wild.
+With an appropriate community process, this could lead to convergence via collections of curated, high-quality collections more quickly than is possible today, where most widely-adopted languages evolve quite slowly.
+%We also anticipate that coding guidelines and tools mandating the use of abstractions that are known to have certain desirable properties will replace designer-mandated enforcement.
 
- We anticipate that coding guidelines mandating the use of abstractions that can be shown to have certain desirable properties will replace language-mandated enforcement opinions
-
- Future work: integrate with something like scalad http://lampwww.epfl.ch/~hmiller/scala2013/resources/pdfs/paper8.pdf. \todo{fix python lang reference citation}
+This paper is presented as a language design paper.
+The theoretical foundations of this work lie in type-level computation.
+We have developed a simplified, type-theoretic formalism of active typechecking and translation, where we prove the safety properties we have outlined here, as well as several additional ones that are only possible to achieve using a typed metalanguage.
+That work is currently in submission at ESOP 2014.
+The work described here is original and extends that mechanism in several directions: Ace supports a rich syntax, makes syntax trees available to extensions, supports a pluggable backend and per-function base semantics.
+It is implemented in an existing language as a library and supports a wide range of eminently practical extensions.
+We hope that it will serve as a useful tool both to researchers and in practice.
+The mechanisms described here can be supported be any language that offers some form of quotation of function ASTs and a reasonably flexible collection of syntactic forms.
+They are particularly useful when the language has a variety of interoperability layers with other languages, like Python does.
 

### 2013-11-20 ace-pldi14/ace-pldi14.tex
pldi14 post-submission cleanup for writing skills req and future revisions

@@
 The \emph{target} of compilation is also user-defined.
-We will show examples of Ace targeting Python as well as OpenCL and CUDA, lower-level languages often used to program graphics hardware.
+We will show examples of Ace targeting Python as well as OpenCL, CUDA and C99, lower-level languages often used to program hardware.
 An active base or type can support multiple \emph{active targets}, which mediate translation of Ace code to code in a target language.
@@
 Listing \ref{map} shows an example of an Ace file.
-As promised, the top level of an Ace file is written directly in Python, requiring no modifications to the language (versions 2.6+ or 3.3+) nor features specific to CPython (so Ace supports alternative implementations like Jython and IronPython).
+As promised, the top level of an Ace file is written directly in Python, requiring no modifications to the language (versions 2.6+ or 3.3+) nor features specific to CPython (so Ace supports alternative implementations like Jython, IronPython and PyPy).
 This choice pays immediate dividends on line 1: Ace's package system is Python's package system, so Python's build tools (e.g.
@@
 Functions containing run-time behavior, like \verb|map|, are governed by a semantics that can differ from Python's (in ways that we will describe below), but they share Python's syntax.
-As a consequence, users of Ace  immediately benefit from an ecosystem of well-developed tools that work with Python syntax, including parsers, code highlighters, editor modes, style checkers and documentation generators.
+As a consequence, users of Ace  immediately benefit also from an ecosystem of well-developed tools that work with Python syntax, including parsers, code highlighters, editor modes, style checkers and documentation generators.
 
@@
 This choice is merely an implementation detail that can be kept private to \verb|clx|, however.
-An Ace value of type \verb|clx.int2| (that is, an actual OpenCL vector) \emph{cannot} be used when a \verb|clx.Cplx(clx.int)| is expected (and attempting to do so will result in a static type error).
-\verb|clx.Cplx| truly extends the type system, it is not a type alias.
+An Ace value of type \verb|clx.int2| (that is, an actual OpenCL vector) \emph{cannot} be used when a \verb|clx.Cplx(clx.int)| is expected (and attempting to do so will result in a static type error): \verb|clx.Cplx| truly extends the type system, it is not a type alias.
 
@@
 Generic functions, according to Sec.
-\ref{genfn}, have not yet had a type assigned to them, so it may seem perplexing that we are nevertheless extracting a type from it.
+\ref{genfn}, have not yet had a type assigned to them, so it may seem perplexing that we are nevertheless assigning a type to \verb|negate|.
 Although a conventional arrow type cannot be assigned to \verb|negate|, we can give it a \emph{singleton type}: a type that simply means ``this expression is the \emph{particular} generic function \verb|negate|''.
 This type could also have been explicitly written as \verb|ace.GenericFnType(listing2.negate)|.
-During typechecking and translation of \verb|map_neg_f32| and \verb|map_neg_ci32|, the call to \verb|f| on line \ref{map}.6 operates by using the types of the provided arguments to compile the generic function that inhabits the singleton type of \verb|f| (\verb|negate| in both of these cases) to a concrete function.
+During typechecking and translation of \verb|map_neg_f32| and \verb|map_neg_ci32|, the call to \verb|f| on line \ref{map}.6 uses the type of the provided argument to compile the generic function that inhabits the singleton type of \verb|f| (\verb|negate| in both of these cases) to a concrete function.
 This is why there are two versions of \verb|negate| in the output in Listing \ref{mapout}.
@@
 The OpenCL language was designed for this workflow, exposing a retargetable compiler and data management  routines as an API, called the \emph{host API} \cite{opencl11}.
-The \verb|pyopencl| library exposes this API and supports basic interoperability with \verb|numpy|, a package for safely manipulating contiguously-allocated numeric arrays in Python \cite{klockner2011pycuda}.
+The \verb|pyopencl| library exposes this API to Python and simplifies interoperation with \verb|numpy|, a popular package for manipulating contiguously-allocated numeric arrays in Python \cite{klockner2011pycuda}.
 
@@
 For developers experienced with the usual OpenCL or CUDA workflow, the fact that this can be accomplished in a total of 6 statements may be surprising.
-This simplicity is possibly largely due to  implicit tracking of types throughout the code.
+This simplicity is made possible by Ace's implicit tracking of types throughout the code.
 First, \verb|numpy| keeps track of the type, shape and order of its arrays.
-The type of \verb|input|, for example, is \verb|numpy.double| by default, its shape is \verb|(1024,1024)| and its order is row-major by default.
-The \verb|pyopencl| library that our mechanism is built atop uses this information to automatically call the underlying OpenCL host API function for transferring byte arrays to the device without requiring the user to calculate the size.
-Our wrapper of \verb|pyopencl| further retains this information in the Python wrappers around the device arrays, \verb|d_input| and \verb|d_output|.
-The Ace active type of such wrappers can be designated to be an instance of \verb|clx.NPArray| parametermized by this metadata.
-This type knows how to typecheck and translate operations like indexing automatically (see Sec.
-\ref{atypes}).
+The type of \verb|input|, for example, is \verb|numpy.float64| by default, its shape is \verb|(1024,1024)| and its order is row-major by default.
+The \verb|pyopencl| library, which the \verb|clx| mechanism is built upon, uses this metadata to automatically call the underlying OpenCL host API function for transferring byte arrays to the device without requiring the user to calculate the size.
+The \verb|clx| wrapper further retains this metadata in \verb|d_input| and \verb|d_output|, the Python wrappers around the allocated device arrays.
+The Ace active type of these wrappers is an instance of \verb|clx.NPArray| parameterized by this metadata.
+This type knows how to typecheck and translate operations, like indexing with a multi-dimensional thread index, automatically.
 
-This allows us to call the generic function \verb|map| directly on these Python data structures (as well as the generic function \verb|negate|) without first requiring an explicit type assignment, like we needed when using \verb|acec| above.
+By extracting the types from the arguments, we can call the generic function \verb|map| without first requiring an explicit type assignment, like we needed when using \verb|acec| above.
 In other words, dynamic types and other metadata can propagate from Python data structures into an Ace generic function as static type information, in the same manner as it propagated \emph{between} generic functions in the previous section.
 In both cases, typechecking and translation of \verb|map| happens the first time a particular type assignment is encountered and cached for subsequent use.
-When called from Python, the generated OpenCL source code is also compiled for the device we selected using the OpenCL host API's compiler infrastructure, and cached.
+When called from Python, the generated OpenCL source code is compiled for the device we selected using the OpenCL retargetable compilation infrastructure, and cached for subsequent calls.
 
 The same program written using the OpenCL C API directly is an order of magnitude longer and significantly more difficult to comprehend.
-OpenCL not support higher-order functions nor is there any way to write \verb|map| in a type-generic manner.
+OpenCL does not support higher-order functions nor is there any way to write \verb|map| in a type-generic or shape-generic manner.
 If we instead use the \verb|pyopencl| library and apply the techniques described in \cite{klockner2011pycuda}, the program is still twice as large and less readable than this code.
-Both the \verb|map| and \verb|negate| functions must be explicitly specialized with the appropriate types using string manipulation techniques.
+Both the \verb|map| and \verb|negate| functions must be explicitly specialized with the appropriate types using string manipulation techniques, and custom shapes and orders can be awkward to handle.
 Higher order functions are still not available, and must also be simulated by string manipulation.
-That approach also does not permit the use any of the language extensions that Ace enables (e.g.
-the type for \verb|numpy| arrays, which ensures that indexing respects ordering; see Sec.
+That approach also does not permit the use any of the language extensions that Ace enables (beyond the useful type for \verb|numpy| arrays just described; see Sec.
 \ref{examples} for more interesting possibilities).
@@
 In a compiler for a monolithic language, there would be a \emph{syntax-directed} protocol governing typechecking and translation.
-In a compiler written in a functional language, for example, one would declare datatypes that captured all forms of types and expressions, and the typechecker would perform exhaustive case analysis over the expression forms, so all the semantics would be implemented in one place.
+In a compiler written in a functional language, for example, one would declare datatypes that captured all forms of types and expressions, and the typechecker would perform exhaustive case analysis over the expression forms.
+That is, all the semantics are implemented in one place.
 The visitor pattern typically used in object-oriented languages implements essentially the same protocol.
-This does not work for an extensible language because new cases and logic need to be added modularly, in a safely composable manner.
+This does not work for an extensible language because new cases and logic need to be added \emph{modularly}, in a safely composable manner.
 
@@
 When encountering a compound term (e.g.
-\verb|e1[e2]|), the compiler defers control over typechecking and translation to the active type of a designated subexpression (e.g.
-\verb|e1|) determined by Ace's fixed \emph{dispatch protocol}.
-Below are examples of these choices.
+\verb|e[e1]|), the compiler defers control over typechecking and translation to the active type of a designated subexpression (e.g.
+\verb|e|) determined by Ace's fixed \emph{dispatch protocol}.
+Below are examples of the choices made in Ace.
 Due to space constraints, we do not show the full protocol.
@@
 We note that error messages are an important and sometimes overlooked facet of {ease-of-use} \cite{marceau2011measuring}.
-A common frustration with using general-purpose abstraction mechanisms to encode an abstraction is that they can produce  verbose and cryptic error messages that reflect the implementation details, rather than the semantics.
-%Active types permit direct encoding of semantics and customization of error messages.
+A common frustration with using general-purpose abstraction mechanisms to encode an abstraction is that they can produce  verbose and cryptic error messages that reflect the implementation details instead of the semantics.
+Ace supports custom error messages.%Active types permit direct encoding of semantics and customization of error messages.
 
@@
 The typechecking rules for this logic is implemented on lines \ref{cplx}.17-\ref{cplx}.29.
-Because arithmetic operations are meant to be symmetric, the dispatch protocol checks both subexpressions for support, favoring the left to ensure that the semantics remain deterministic.
-In either position, the implementation begins by recursively assigning a type to the other operand in the current context via the \verb|context.type| method (line \ref{cplx}.24).
-If supported, it applies the C99 rules for arithmetic operations to determine the resulting type (not shown).
+Arithmetic operations are usually symmetric, so the dispatch protocol checks the types of both subexpressions for support.
+To ensure that the semantics remain deterministic in the case that both types support the binary operation, Ace asks the left first (via \verb|type_BinOp_left|), asking the right (via \verb|type_BinOp_right|) only if the left indicates an error.
+In either position, our implementation begins by recursively assigning a type to the other operand in the current context via the \verb|context.type| method (line \ref{cplx}.24).
+If supported, it applies the C99 rules for arithmetic operations to determine the resulting type (via \verb|c99_binop_t|, not shown).
 
@@
 Because \verb|clx.Cplx| evaluates not to an active type, but to a class, this form is assigned a type by handing control to the class object via the \emph{class method} \verb|type_New|.
-It operates as expected, extracting the types of the two arguments to construct an appropriate complex number type (lines \ref{cplx}.50-\ref{cplx}.57), raising a type error if the arguments cannot be promoted to a common type according to the rules of C99 or if two arguments have not been provided (an exercise for the reader: modify this method to also allow a single argument for when the imaginary part is 0).
+It operates as expected, extracting the types of the two arguments to construct an appropriate complex number type (lines \ref{cplx}.50-\ref{cplx}.57), raising a type error if the arguments cannot be promoted to a common type according to the rules of C99 or if two arguments were not provided.% (an exercise for the reader: modify this method to also allow a single argument for when the imaginary part is 0).
 
@@
 Once typechecking a method is complete, the compiler enters the translation phase, where terms in the target language are generated from Ace terms.
-Terms in the target language are generated by calling methods of the \emph{active targets}.
+Terms in the target language are generated by calling methods of the \emph{active target} governing the function being compiled.
 The translation phase operates similarly to typechecking, using the dispatch protocol to invoke methods named \verb|trans_|$X$.
-These methods have access to the context and node just as during typechecking, as well as the target.
+These methods have access to the context and node just as during typechecking, as well as the active target (named \verb|target| here).
 
 As seen in Listing \ref{mapout}, we are implementing complex numbers internally using OpenCL vector types, like \verb|int2|.
-Let us look first at \verb|trans_New| on lines \ref{cplx}.54-\ref{cplx}.60, where new complex numbers are translated to vector literals by invoking \verb|VecLit| from the active target, which internally will generate the necessary OpenCL string for that form.
-The \verb|trans_type| method of the \verb|ace.ActiveType| is used to associated a type in the target language with an active type.
-For it to be possible to reason compositionally about the correctness of compilation, all complex numbers must translate to terms in the target language that have this target type.
-Ace supports a mode where this \emph{representational consistency} is dynamically checked during compilation (this requires that the active target know how to assign types to terms in the target language, which can be done only for our OpenCL target as of this writing).
+Let us look first at \verb|trans_New| on lines \ref{cplx}.54-\ref{cplx}.60, where new complex numbers are translated to vector literals by invoking \verb|target.VecLit|.
+This will ultimately  generate the necessary OpenCL code, as a string, to complete compilation (these strings are not directly manipulated by extensions, however, to avoid problems with, e.g.
+precedence).
+For it to be possible to reason compositionally about the correctness of compilation, all complex numbers must translate to terms in the target language that have a consistent target type.
+The \verb|trans_type| method of the \verb|ace.ActiveType| associates a type in the target language, here a vector type like \verb|int2|, with the active type.
+Ace supports a mode where this \emph{representational consistency} is dynamically checked during compilation (requiring that the active target know how to assign types to terms in the target language, which can be done  for our OpenCL target as of this writing).
 
-The translation methods for attributes and binary operations proceed in a straightforward manner.
+The translation methods for attributes (line \ref{cplx}.12) and binary operations  (line \ref{cplx}.31) proceed in a straightforward manner.
 The context provides a method, \verb|trans|, for recursively determining the translation of subexpressions as needed.
@@
 For example, the implementation of \verb|trans_Attribute| assumes that if \verb|node.attr| is not \verb|'ni'| then it must have been \verb|'i'| on line \ref{cplx}.14, consistent with the implementation of \verb|type_Attribute| above it.
-Typechecking and translation are separated into two methods to emphasize that typechecking is not target-dependent and to allow for typing rules that require generating types for hypothetical terms, where the overhead of the translation might be better avoided.
+Typechecking and translation are separated into two methods to emphasize that typechecking is not target-dependent, and to allow for more advanced uses, like type refinements and hypothetical typing judgements, that we do not describe here.
 
@@
 A base is handed control over typechecking of statements and expressions in the same way as active types: via \verb|type_|$X$ and \verb|trans_|$X$ methods.
-It also initializes the context and governs the semantics of variables.
-Each function can have a distinct base semantics.
+Each function can have a different base.
 
@@
 Similarly, for some applications, avoiding accidental overflows is more important than performance.
-Arbitrary-precision integers can be implemented as an active type (not shown) and the base generates the constructor call form, described above, around integers to support this.
+Infinite-precision integers can be implemented as an active type (not shown) and the base can be asked to use it for numeric literals.
+In all cases, this occurs by inserting a constructor call form, as described above.% the constructor call form, described above, around integers to support this.
 
-Another role that the active base can play is to provide operators that do not need to be imported.
-The semantics of \verb|get_global_id| and \verb|printf|, for example, are provided by \verb|clx.base|.
-In fact, the base provides a richer semantics for them than the underlying implementation, as we will show in the following section.
-It is not strictly necessary that the base provide these -- they could have been imported from \verb|clx| and used by their qualified name.
-A base which does not provide them without import can be generated by passing the \verb|primitives=False| flag in.
+The base also initializes the context and governs the semantics of variables and assignment.
+This can also permit it to allow for the use of ``built-in'' operators that were not explicitly imported.
+The semantics of \verb|get_global_id| and \verb|printf| are provided by \verb|clx.base|.
+In fact, the base provides extended versions of them (the former permits multi-dimensional global IDs, the latter supports typechecking format strings).A base which does not provide them as built-ins (requiring that they be imported explicitly) can be generated by passing the \verb|primitives=False| flag.
 
-\subsection{Active Targets}\label{atargets}
-An active target, which we show being used in the translation logic described above, is an instance of \verb|ace.ActiveTarget|.
-It has two primary roles in Ace: 1) it provides an API for code generation to a target language, providing term and type constructors for use during translation (and, to support representational consistency checks, a type system implementation); 2) it provides a bridge to the underlying interoperability API when a generic or concrete function is launched directly from Python, as described in Sec.
+\subsection{Active Targets and Cross-Compilation}\label{atargets}
+An active target, which we describe being used above, is an instance of \verb|ace.ActiveTarget|.
+It has two primary roles in Ace: 1) providing an API for code generation to a target language, providing term and type constructors for use during translation (and, optionally, a type system implementation to support representational consistency checks); 2) implementing the interoperability layer to support calling of  generic or concrete function directly from Python, as described in Sec.
 \ref{implicit}.
 
-An active type or active base can support multiple targets if desired by simply examining it during translation (e.g.
-by performing capability checks using Python's \verb|hasattr| function) and providing alternative code paths for different combinations of capabilities.
-Alternatively, an active backend can implement the interfaces available in a different back end and generate code for them in a different way.
-Our implementation of OpenCL's type system as a collection of active types and an active base does not explicitly support cross-compilation to a CUDA target (though it would be possible to modify it to do so), but our CUDA target does provide partial support for the OpenCL code generation API.
+An active type or base can support multiple active targets by simply examining \verb|target| during translation (e.g.
+by performing capability checks using Python's \verb|hasattr| function).
+Alternatively, the active target can simulate the interface of a different target but generate code differently.
+For example, although our implementation of OpenCL's type system as a collection of active types and an active base relies on a target that supports OpenCL terms and types, our \verb|CUDA| and \verb|C99| targets provide partial support for the same term and type forms, in order to support cross-compilation.
+We will show the latter being used in Sec.
+\ref{examples}.
 
@@
 
-\subsection{Functional Datatypes}
+\subsection{Recursive Labeled Sums}
 \begin{codelisting}
@@
 \end{codelisting}
-Listing \ref{datatypest} shows an example of the use of statically-typed functional datatypes alongside the Python implementation just described.
-It shows two syntactic conveniences that were not mentioned previously: 1) if no target is provided to \verb|ace.fn|, the base can provide a default target (here, \verb|ace.python.python|; 2) A concrete function can be generated immediately by providing a type assignment to \verb|ace.fn|.
+Listing \ref{datatypest} shows an example of the use of statically-typed functional datatypes (a.k.a.
+recursive labeled sum types) \todo{change code to take a lambda} together with the Python implementation just described.
+It shows two syntactic conveniences that were not mentioned previously: 1) if no target is provided to \verb|ace.fn|, the base can provide a default target (here, \verb|ace.python.python|); 2) a concrete function can be generated immediately by providing a type assignment after \verb|ace.fn| using braces.
 Listing \ref{datatypest} generates Listing \ref{datatypes}.
@@
 
-\subsection{Structs, Unions, Records and Objects}
-Data structures like structs, unions, records and objects can all be implemented as Ace type families  parameterized by a dictionary mapping field names to types.
+\subsection{Product Types}
+Data structures like structs, unions, records and objects\todo{mention that these are product types} can all be implemented as Ace type families  parameterized by a dictionary mapping field names to types.
 Each family differs slightly in the semantics of the operations available.

### 2013-11-20 ace-pldi14/listing7.py

@@ -6,4 +6,4 @@ scale = _fn(ast.parse("""def scale(x, s):
 	return x * s""")
 
 negate = _fn(astx.specialize(scale.ast, "negate", 
-	s=ast.parse("-1"))
+	s=astx.parse_exp("-1"))

### 2013-11-20 ace-pldi14/ace-pldi14.tex
Revisions of PLDI14 for speaking skills

@@
 %type-level computation, typed compilation
-\section{Introduction}
+\section{Introduction}\label{intro}
 Asking programmers to import a new library is far more practical than asking them to  adopt a new programming language.
@@
 Listing \ref{datatypest} shows an example of the use of statically-typed functional datatypes (a.k.a.
-recursive labeled sum types) \todo{change code to take a lambda} together with the Python implementation just described.
+recursive labeled sum types) together with the Python implementation just described.
 It shows two syntactic conveniences that were not mentioned previously: 1) if no target is provided to \verb|ace.fn|, the base can provide a default target (here, \verb|ace.python.python|); 2) a concrete function can be generated immediately by providing a type assignment after \verb|ace.fn| using braces.
@@
 This type implemented by the active type family \verb|fp.Datatype|.
-Case names and their types are provided programmatically on lines \ref{datatypest}.5-\ref{datatypest}.10.
+A name for the datatype and the case names and types are provided programmatically on lines \ref{datatypest}.3-\ref{datatypest}.8.
+To support recursive datatypes, the case names are enclosed within a lambda term that will be passed a reference to the datatype itself.
 These lines also show two more active type families: units (the type containing just one value), and immutable pairs.
@@
 The patterns to the left of the colons are not treated as expressions (that is, a type and translation is never recursively assigned to them).
-The active type implements the standard algorithm for ensuring that the cases cover all possibilities and are not redundant.
-If the final case were omitted, for example, the algorithm would statically indicate an error, just as in statically-typed functional languages like ML and Haskell.
+Instead, the active type implements the standard algorithm for ensuring that the cases cover all possibilities and are not redundant \cite{pfpl}.
+If the final case were omitted, for example, the algorithm would statically indicate an error, just as in statically-typed functional languages like ML.
 
@@
 \subsection{Product Types}
-Data structures like structs, unions, records and objects\todo{mention that these are product types} can all be implemented as Ace type families  parameterized by a dictionary mapping field names to types.
+Product types (types that group heterogeneous values together) like structs, unions, tuples, records and objects can all be implemented as Ace type families  parameterized by a dictionary mapping field names (indices, in the case of tuples) to types.
 Each family differs slightly in the semantics of the operations available.
@@
 Listing \ref{ooclxfppy} shows an example of each of these.
-The prototypic object system delegates to a backing record if a field is not available in the struct.
+The prototypic object system delegates to a backing record if a field is not available in the foreground.
 This example also demonstrates cross-compilation to C99, supported for a subset of operations in \verb|clx|.
 All three abstractions are implemented as simple \verb|struct|s, as can be seen in Listing \ref{ooclxfpc}.
-This example also demonstrates the use of argument annotations to generate concrete functions (line \ref{ooclxfppy}.9), which is a feature only available when running Ace in Python 3.
+This example further demonstrates the use of argument annotations to generate concrete functions (line \ref{ooclxfppy}.9), which is a feature only available when using Python 3.3+.
 
@@
 Many distributed programming abstractions can be understood as implementing object models with complex forms of dynamic dispatch.
-For example, in Charm++, dynamic dispatch involves message passing over a dynamically load-balanced network \cite{kale1993charm++}.
+For example, in Charm++, dynamic dispatch involves message passing over a dynamically load-balanced network \cite{kale2009charm++}.
 While Charm++ is a sophisticated system, and we do not anticipate that implementing it as an Ace extension would be easy, it is possible to do so by targeting a system like Adaptive MPI, which exposes the Charm++ runtime system \cite{kale2009charm++}.
@@
 
-A number of recent languages designed for distributed programming on clusters use a partitioned global address space model, including UPC \cite{upc}, Chapel \cite{chapel} and others.
+A number of recent languages designed for distributed programming on clusters use a partitioned global address space model, e.g.
+Chapel \cite{chapel}.
 These languages provide first-class support for accessing data transparently across a massively parallel cluster.
@@
 \end{longtable}
-\caption{Comparison to related approaches to language-internal extensibility.}
+\caption{Comparison to related approaches to language-internal extensibility.}\label{relatedtable}
 \twocolumn
@@
 
-Libraries that contain compile-time logic have been called {\it active libraries} in prior proposals \cite{activelibraries}.
-A number of  projects, such as Blitz++, have taken advantage of the C++ preprocessor and template-based metaprogramming system to implement domain-specific optimizations \cite{veldhuizen2000blitz++}.
+Some previous work has considered embedding statically-typed languages within other languages, including dynamically-typed languages.
+For example, Terra embeds a low-level statically-typed language within Lua \cite{terra}.
+These have not been fundamentally extensible, and generally require modifying the host language externally, leading to the ``bootstrapping'' problem of Sec.
+\ref{intro}.
+
+Libraries containing compile-time logic have previously been called {\it active libraries} \cite{activelibraries}.
+A number of  projects, such as Blitz++, have taken advantage of C++ template metaprogramming to implement domain-specific optimizations \cite{veldhuizen2000blitz++}.
+Others have chosen custom metalanguages (e.g.
+Xroma \cite{activelibraries}).
 In Ace, we replace these brittle mini-languages with a general-purpose language, Python, and significantly expand the notion of active libraries by consideration of types, base semantics and target languages as values (in this case, objects) in the compile-time metalanguage.
@@
 In Ace, generic functions are implicitly polymorphic and can be called with arguments of {\it any type that supports the operations used by the function}.
-This approach is related to structural polymorphism \cite{malayeri2009structural}.
+This is related to structural polymorphism \cite{malayeri2009structural}.
 Structural types make explicit the requirements on a function, unlike generic functions, which are only given singleton types because the requirements can be defined arbitrarily by extensions.
@@
 However, our strategy is a {\it compile-time} protocol where the typechecking and translation semantics are implemented for different operators.
-Our dispatch protocol was designed to be conceptually similar to Python's operator overloading protocol \cite{python}.
+%Our dispatch protocol was designed to be conceptually similar to Python's operator overloading protocol \cite{python}.
 
 There are several other language-internal extension mechanisms that have been described in the literature.
-These differ from our mechanism based primarily on active types in one of the following ways, summarized in the table in Figure 1:
+These differ from our mechanism in one of the following ways, summarized in Fig.
+\ref{relatedtable}:
 \begin{itemize}
@@
 \item Our mechanism reuses an existing language's syntax, rather than offering syntax extension support.
-As we saw in, for example, Listing \ref{datatypest}, this can still be reasonably natural.
-However, some mechanisms provide true syntax extensions.
+As we saw in, for example, Listing \ref{datatypest}, this can still be reasonably natural, and allows the language to benefit from a variety of existing tools.
 \item Our mechanism permits true type system extensions.
-New types are not merely aliases, nor must their rules be admissible in some base type system.
-Term rewriting and macro systems do not handle types at all.
+New types are not merely aliases, nor must their rules be admissible in some base type system.% Term rewriting and macro systems do not handle types at all.
 \item As described in Sec.
@@
 Techniques that allow global extensions to the syntax or semantics (e.g.
-SugarJ or rule injection systems like Xroma or Typed Racket) allow extensions so much control that they can interfere with one another or make reasoning about code difficult.
-\item Our mechanism allows users to control the target language of compilation from within libraries. Many other mechanisms are based fundamentally on rewriting mechanisms.
+SugarJ or rule injection systems like Xroma or Typed Racket) allow extensions so much control that they can interfere with one another or make reasoning  difficult.
+\item Our mechanism allows users to control the target language of compilation from within libraries. Many other mechanisms are based fundamentally on term-to-term rewriting.
 \end{itemize}
@@
 When the mechanisms available in an existing language prove insufficient, researchers and domain experts often design a new language.
-A number of tools have been developed to assist with this task, including compiler generators, language workbenches and domain-specific language frameworks (cf \cite{fowler2010domain}).
+A number of tools have been developed to assist with this task, including compiler generators, language workbenches and domain-specific language frameworks (cf \cite{erdweg2013state}).
 Extensible compilers can be considered a form of language framework as well due to the interoperability issues that relying on a compiler-specific extensions can introduce.
-It is difficult or impossible for these language-external approaches to achieve interoperability, because languages are not aware of each other's semantics.
+It is difficult or impossible for these language-external approaches to achieve interoperability, because languages are not aware of each other's semantics and merging languages is not guaranteed to be sound.
 
@@
 Debuggers and other tools that rely not just on Python's syntax but also its semantics cannot be used directly, so if code generation introduces significant complexity that leads to bugs, this can be an issue.
-The OpenCL library we have implemented is a reasonably straightforward internalization of OpenCL itself, however, so debugging has not been a problem thusfar.
 We believe that active types can be used to control debugging, and plan to explore this in the future.
 We have also not yet evaluated the feasibility of implementing more advanced type systems (e.g.
-linear type systems, dependent type systems, flow-dependent type systems) using Ace.
+linear, dependent or flow-dependent type systems) using Ace.
 
@@
 Instead, we argue that the potential for abuse must be balanced with the possibilities made available by a vibrant ecosystem of competing statically-typed abstractions that can be developed and deployed as libraries, and thus more easily evaluated in the wild.
-With an appropriate community process, this could lead to convergence via collections of curated, high-quality collections more quickly than is possible today, where most widely-adopted languages evolve quite slowly.
+With an appropriate community process, this could lead to a convergence toward stable collections of curated, high-quality collections more quickly than is possible today.
 %We also anticipate that coding guidelines and tools mandating the use of abstractions that are known to have certain desirable properties will replace designer-mandated enforcement.
@@
 The theoretical foundations of this work lie in type-level computation.
-We have developed a simplified, type-theoretic formalism of active typechecking and translation, where we prove the safety properties we have outlined here, as well as several additional ones that are only possible to achieve using a typed metalanguage.
-That work is currently in submission at ESOP 2014.
-The work described here is original and extends that mechanism in several directions: Ace supports a rich syntax, makes syntax trees available to extensions, supports a pluggable backend and per-function base semantics.
-It is implemented in an existing language as a library and supports a wide range of eminently practical extensions.
-We hope that it will serve as a useful tool both to researchers and in practice.
-The mechanisms described here can be supported be any language that offers some form of quotation of function ASTs and a reasonably flexible collection of syntactic forms.
-They are particularly useful when the language has a variety of interoperability layers with other languages, like Python does.
+We have developed a simplified, type-theoretic formalism of active typechecking and translation, where we prove the safety properties we have outlined here, as well as several additional ones that are only possible to achieve using a typed metalanguage (currently in submission at ESOP 2014).
+The work described here  extends that mechanism in several directions: Ace supports a rich syntax, makes syntax trees available to extensions, supports a pluggable backend and per-function base semantics.
+It is implemented in an existing language and supports a wider range of practical extensions.
+
+The mechanisms described here can be implemented within any language that offers some form of quotation of function ASTs and a reasonably flexible collection of syntactic forms and basic reflection facilities.
+It is particularly useful when the language supports interoperability layers with a number of target languages.
 

### 2013-11-20 ace-pldi14/datatypes_t.py

@@ -1,20 +1,17 @@
 import ace, ace.python as py, examples.fp as fp
 
-def Tree(name, a):
-  t = fp.Datatype(name)
-  t.cases = {
+Tree = lambda name, a: fp.Datatype(name, 
+  lambda tree: {
     'Empty': fp.unit,
     'Leaf': a,
-    'Node': fp.tuple(t, t)
-  }
-  t.close()
-  return t
+    'Node': fp.tuple(tree, tree)
+  })
 
 DT = Tree('DT', py.dyn)
 
 @ace.fn(py.Base(trailing_return=True))[[DT]]
-def depth_gt_2(t):
-  t.case({
+def depth_gt_2(x):
+  x.case({
     DT.Node(DT.Node(_), _): True,
     DT.Node(_, DT.Node(_)): True,
     _: False

### 2013-11-21 ace-pldi14/ace-pldi14.tex
Added Rust reference to PLDI14

@@
 \begin{figure*}
+\vspace{-10pt}
 \onecolumn
@@
 \twocolumn
+\vspace{-10pt}
 \end{figure*}
@@
 For example, Terra embeds a low-level statically-typed language within Lua \cite{terra}.
-These have not been fundamentally extensible, and generally require modifying the host language externally, leading to the ``bootstrapping'' problem of Sec.
+Our primary example showed how to support low-level GPU programming from a high-level language.
+Other languages have supported similar workflows, e.g.
+Rust \cite{rustgpu}.
+These have not been fundamentally extensible and suffer from the ``bootstrapping'' problem described in Sec.
 \ref{intro}.
@@
 This is related to structural polymorphism \cite{malayeri2009structural}.
-Structural types make explicit the requirements on a function, unlike generic functions, which are only given singleton types because the requirements can be defined arbitrarily by extensions.
+Structural types make explicit the structure required of an argument, unlike generic functions, which are only given singleton types because the structure may depend on the semantics of an active type.
 Structural typing can be compared to the more \emph{ad hoc} approach taken by dynamically-typed languages like Python itself, sometimes called ``duck typing''.
 It is also comparable to the C++ template system, as discussed previously.
-Debugging type errors can be more difficult when there are many nested generic functions that were implicitly given type assignments.
-Type debuggers like \verb|scalad| could help make this easier \cite{scalad}.
 
@@
 Debuggers and other tools that rely not just on Python's syntax but also its semantics cannot be used directly, so if code generation introduces significant complexity that leads to bugs, this can be an issue.
+Debugging type errors can also be difficult when there are many nested generic functions that were implicitly given type assignments.
+Type debuggers like \verb|scalad| could help make this easier \cite{scalad}.
 We believe that active types can be used to control debugging, and plan to explore this in the future.

### 2013-11-21 ace-pldi14/ace-pldi14.tex
Added mbeddr reference to PLDI14

@@
 \verb|clx.NPArray|).
-Ace was first conceived to answer the question: \emph{can we build a statically-typed language with the semantics of a C but the syntax and ease-of-use of a Python?} We submit that these examples (along with a large-scale simulator framework built using them, which we do not have space to detail here) have answered this question in the affirmative.
+Ace was first conceived to answer the question: \emph{can we build a statically-typed language with the semantics of a C but the syntax and ease-of-use of a Python?} We submit that the work described above has answered this question in the affirmative.
 
@@
 
-Rule Injection && Qi, Typed Racket \cite{TypedScheme2008}, Xroma \cite{activelibraries} && some && \Circle && \CIRCLE && \Circle && \Circle \\
+Rule Injection && Racket \cite{TypedScheme2008}, Xroma \cite{activelibraries}, mbeddr \cite{mbeddr} &&  \cite{TypedScheme2008} && \cite{mbeddr} && \CIRCLE && \Circle && \Circle \\
 
@@
 Others have chosen custom metalanguages (e.g.
-Xroma \cite{activelibraries}).
+Xroma \cite{activelibraries}, mbeddr \cite{mbeddr}).
 In Ace, we replace these brittle mini-languages with a general-purpose language, Python, and significantly expand the notion of active libraries by consideration of types, base semantics and target languages as values (in this case, objects) in the compile-time metalanguage.
@@
 \begin{itemize}
-\item Our mechanism is itself implemented as a library, rather than as a language extension.
+\item Our mechanism is itself implemented as a library, rather than as a language extension. Only typed LISPs are similar.
 \item Our mechanism reuses an existing language's syntax, rather than offering syntax extension support.
@@
 Techniques that allow global extensions to the syntax or semantics (e.g.
-SugarJ or rule injection systems like Xroma or Typed Racket) allow extensions so much control that they can interfere with one another or make reasoning  difficult.
+SugarJ or rule injection systems like Xroma, Typed Racket (and other typed LISPs) and mbeddr) allow extensions so much control that they can interfere.
 \item Our mechanism allows users to control the target language of compilation from within libraries. Many other mechanisms are based fundamentally on term-to-term rewriting.

### 2013-11-22 sciunit-icse14/sciunit-icse14.tex
Added initial dump of SciUnit paper from Frontiers

%% bare_conf.tex
%% V1.3
%% 2007/01/11
%% by Michael Shell
%% See:
%% http://www.michaelshell.org/
%% for current contact information.
%%
%% This is a skeleton file demonstrating the use of IEEEtran.cls
%% (requires IEEEtran.cls version 1.7 or later) with an IEEE conference paper.
%%
%% Support sites:
%% http://www.michaelshell.org/tex/ieeetran/
%% http://www.ctan.org/tex-archive/macros/latex/contrib/IEEEtran/
%% and
%% http://www.ieee.org/

%%*************************************************************************
%% Legal Notice:
%% This code is offered as-is without any warranty either expressed or
%% implied; without even the implied warranty of MERCHANTABILITY or
%% FITNESS FOR A PARTICULAR PURPOSE! 
%% User assumes all risk.
%% In no event shall IEEE or any contributor to this code be liable for
%% any damages or losses, including, but not limited to, incidental,
%% consequential, or any other damages, resulting from the use or misuse
%% of any information contained here.
%%
%% All comments are the opinions of their respective authors and are not
%% necessarily endorsed by the IEEE.
%%
%% This work is distributed under the LaTeX Project Public License (LPPL)
%% ( http://www.latex-project.org/ ) version 1.3, and may be freely used,
%% distributed and modified. A copy of the LPPL, version 1.3, is included
%% in the base LaTeX documentation of all distributions of LaTeX released
%% 2003/12/01 or later.
%% Retain all contribution notices and credits.
%% ** Modified files should be clearly indicated as such, including  **
%% ** renaming them and changing author support contact information. **
%%
%% File list of work: IEEEtran.cls, IEEEtran_HOWTO.pdf, bare_adv.tex,
%%                    bare_conf.tex, bare_jrnl.tex, bare_jrnl_compsoc.tex
%%*************************************************************************
\documentclass[9pt]{sig-alternate}

%\usepackage{biblatex}
%\usepackage{amsmath}
%\usepackage{amssymb}
%\usepackage{times}
\usepackage{hyperref}
%\usepackage{graphicx}
\usepackage{wrapfig}
\usepackage{listings}
\usepackage[usenames,dvipsnames]{xcolor}
%\usepackage{setspace}
\usepackage{python_syntax}
%\usepackage{bibentry}
\usepackage{todonotes}
\usepackage{url}
\usepackage{lineno}
%\usepackage{natbib}
%\usepackage{caption}
\usepackage[english]{babel} 
\def\UrlFont{\em} % Italicize all URLs.  
\lstset{
	basicstyle=\small\ttfamily,
	breaklines=true,
	language=Python,
	}
\let\verbx\lstinline
\newcommand{\bfhead}[1]{\noindent \textbf{#1:}}

%\usepackage{alltt}
%\renewcommand{\ttdefault}{txtt}
%%\usepackage{color}
%%\usepackage[usenames,dvipsnames,svgnames,table]{xcolor}
%%\definecolor{mauve}{rgb}{0.58,0,0.82}
%\definecolor{light-gray}{gray}{0.75}
%\usepackage{listings}
%\lstset{
%  language=Python,
%  showstringspaces=false,
%  formfeed=\newpage,
%  tabsize=4,
%  commentstyle=\itshape,
%  basicstyle=\ttfamily\scriptsize,
%  morekeywords={lambda, self, assert, as},
%  numbers=left,
%  numberstyle=\scriptsize\color{light-gray}\textsf,
%  xleftmargin=2em,
%  stringstyle=\color{mauve}
%}
%\lstdefinestyle{Bash}{
%    language={}, 
%    moredelim=**[is][\color{blue}\bf\ttfamily]{`}{`},
%}
%\lstdefinestyle{OpenCL}{
%	language=C++,
%	morekeywords={kernel, __kernel, global, __global, size_t, get_global_id, sin}
%}
%
%\usepackage{float}
%\floatstyle{ruled}
%\newfloat{codelisting}{tp}{lop}
%\floatname{codelisting}{Listing}

% *** GRAPHICS RELATED PACKAGES ***
%
% *** MATH PACKAGES ***
%
%\usepackage[cmex10]{amsmath}
% A popular package from the American Mathematical Society that provides
% many useful and powerful commands for dealing with mathematics. If using
% it, be sure to load this package with the cmex10 option to ensure that
% only type 1 fonts will utilized at all point sizes. Without this option,
% it is possible that some math symbols, particularly those within
% footnotes, will be rendered in bitmap form which will result in a
% document that can not be IEEE Xplore compliant!
%
% Also, note that the amsmath package sets \interdisplaylinepenalty to 10000
% thus preventing page breaks from occurring within multiline equations. Use:
%\interdisplaylinepenalty=2500
% after loading amsmath to restore such page breaks as IEEEtran.cls normally
% does. amsmath.sty is already installed on most LaTeX systems. The latest
% version and documentation can be obtained at:
% http://www.ctan.org/tex-archive/macros/latex/required/amslatex/math/

% *** SPECIALIZED LIST PACKAGES ***
%
%\usepackage{algorithmic}
% algorithmic.sty was written by Peter Williams and Rogerio Brito.
% This package provides an algorithmic environment fo describing algorithms.
% You can use the algorithmic environment in-text or within a figure
% environment to provide for a floating algorithm. Do NOT use the algorithm
% floating environment provided by algorithm.sty (by the same authors) or
% algorithm2e.sty (by Christophe Fiorio) as IEEE does not use dedicated
% algorithm float types and packages that provide these will not provide
% correct IEEE style captions. The latest version and documentation of
% algorithmic.sty can be obtained at:
% http://www.ctan.org/tex-archive/macros/latex/contrib/algorithms/
% There is also a support site at:
% http://algorithms.berlios.de/index.html
% Also of interest may be the (relatively newer and more customizable)
% algorithmicx.sty package by Szasz Janos:
% http://www.ctan.org/tex-archive/macros/latex/contrib/algorithmicx/

% *** ALIGNMENT PACKAGES ***
%
%\usepackage{array}
% Frank Mittelbach's and David Carlisle's array.sty patches and improves
% the standard LaTeX2e array and tabular environments to provide better
% appearance and additional user controls. As the default LaTeX2e table
% generation code is lacking to the point of almost being broken with
% respect to the quality of the end results, all users are strongly
% advised to use an enhanced (at the very least that provided by array.sty)
% set of table tools. array.sty is already installed on most systems. The
% latest version and documentation can be obtained at:
% http://www.ctan.org/tex-archive/macros/latex/required/tools/

%\usepackage{mdwmath}
%\usepackage{mdwtab}
% Also highly recommended is Mark Wooding's extremely powerful MDW tools,
% especially mdwmath.sty and mdwtab.sty which are used to format equations
% and tables, respectively. The MDWtools set is already installed on most
% LaTeX systems. The lastest version and documentation is available at:
% http://www.ctan.org/tex-archive/macros/latex/contrib/mdwtools/

% IEEEtran contains the IEEEeqnarray family of commands that can be used to
% generate multiline equations as well as matrices, tables, etc., of high
% quality.

%\usepackage{eqparbox}
% Also of notable interest is Scott Pakin's eqparbox package for creating
% (automatically sized) equal width boxes - aka "natural width parboxes".
% Available at:
% http://www.ctan.org/tex-archive/macros/latex/contrib/eqparbox/

% *** SUBFIGURE PACKAGES ***
%\usepackage[tight,footnotesize]{subfigure}
% subfigure.sty was written by Steven Douglas Cochran. This package makes it
% easy to put subfigures in your figures. e.g., "Figure 1a and 1b". For IEEE
% work, it is a good idea to load it with the tight package option to reduce
% the amount of white space around the subfigures. subfigure.sty is already
% installed on most LaTeX systems. The latest version and documentation can
% be obtained at:
% http://www.ctan.org/tex-archive/obsolete/macros/latex/contrib/subfigure/
% subfigure.sty has been superceeded by subfig.sty.

%\usepackage[caption=false]{caption}
%\usepackage[font=footnotesize]{subfig}
% subfig.sty, also written by Steven Douglas Cochran, is the modern
% replacement for subfigure.sty. However, subfig.sty requires and
% automatically loads Axel Sommerfeldt's caption.sty which will override
% IEEEtran.cls handling of captions and this will result in nonIEEE style
% figure/table captions. To prevent this problem, be sure and preload
% caption.sty with its "caption=false" package option. This is will preserve
% IEEEtran.cls handing of captions. Version 1.3 (2005/06/28) and later 
% (recommended due to many improvements over 1.2) of subfig.sty supports
% the caption=false option directly:
%\usepackage[caption=false,font=footnotesize]{subfig}
%
% The latest version and documentation can be obtained at:
% http://www.ctan.org/tex-archive/macros/latex/contrib/subfig/
% The latest version and documentation of caption.sty can be obtained at:
% http://www.ctan.org/tex-archive/macros/latex/contrib/caption/

% *** FLOAT PACKAGES ***
%
%\usepackage{fixltx2e}
% fixltx2e, the successor to the earlier fix2col.sty, was written by
% Frank Mittelbach and David Carlisle. This package corrects a few problems
% in the LaTeX2e kernel, the most notable of which is that in current
% LaTeX2e releases, the ordering of single and double column floats is not
% guaranteed to be preserved. Thus, an unpatched LaTeX2e can allow a
% single column figure to be placed prior to an earlier double column
% figure. The latest version and documentation can be found at:
% http://www.ctan.org/tex-archive/macros/latex/base/

%\usepackage{stfloats}
% stfloats.sty was written by Sigitas Tolusis. This package gives LaTeX2e
% the ability to do double column floats at the bottom of the page as well
% as the top. (e.g., "\begin{figure*}[!b]" is not normally possible in
% LaTeX2e). It also provides a command:
%\fnbelowfloat
% to enable the placement of footnotes below bottom floats (the standard
% LaTeX2e kernel puts them above bottom floats). This is an invasive package
% which rewrites many portions of the LaTeX2e float routines. It may not work
% with other packages that modify the LaTeX2e float routines. The latest
% version and documentation can be obtained at:
% http://www.ctan.org/tex-archive/macros/latex/contrib/sttools/
% Documentation is contained in the stfloats.sty comments as well as in the
% presfull.pdf file. Do not use the stfloats baselinefloat ability as IEEE
% does not allow \baselineskip to stretch. Authors submitting work to the
% IEEE should note that IEEE rarely uses double column equations and
% that authors should try to avoid such use. Do not be tempted to use the
% cuted.sty or midfloat.sty packages (also by Sigitas Tolusis) as IEEE does
% not format its papers in such ways.

% *** PDF, URL AND HYPERLINK PACKAGES ***
%
% url.sty was written by Donald Arseneau. It provides better support for
% handling and breaking URLs. url.sty is already installed on most LaTeX
% systems. The latest version can be obtained at:
% http://www.ctan.org/tex-archive/macros/latex/contrib/misc/
% Read the url.sty source comments for usage information. Basically,
% \url{my_url_here}.

%\usepackage{placeins}

% *** Do not adjust lengths that control margins, column widths, etc. ***
% *** Do not use packages that alter fonts (such as pslatex).         ***
% There should be no need to do such things with IEEEtran.cls V1.6 and later.
% (Unless specifically asked to do so by the journal or conference you plan
% to submit to, of course. )

% correct bad hyphenation here
\hyphenation{op-tical net-works semi-conduc-tor}

\begin{document}
%
% paper title
% can use linebreaks \\ within to get better formatting as desired
\title{Collaborative Test-Driven Scientific Model Validation}

% author names and affiliations
% use a multiple column layout for up to two different
% affiliations
\numberofauthors{2}
\author{\alignauthor
Cyrus Omar, Jonathan Aldrich\\\affaddr{Carnegie Mellon University, Pittsburgh, PA}\\
       \email{\{comar,aldrich\}@cs.cmu.edu}
       \alignauthor Rick Gerkin\\
       \affaddr{Arizona State University}\\{rgerkin@asu.edu}
}

% conference papers do not typically use \thanks and this command
% is locked out in conference mode. If really needed, such as for
% the acknowledgment of grants, issue a \IEEEoverridecommandlockouts
% after \documentclass

% for over three affiliations, or if they all won't fit within the width
% of the page, use this alternative format:
% 
%\author{\IEEEauthorblockN{Michael Shell\IEEEauthorrefmark{1},
%Homer Simpson\IEEEauthorrefmark{2},
%James Kirk\IEEEauthorrefmark{3}, 
%Montgomery Scott\IEEEauthorrefmark{3} and
%Eldon Tyrell\IEEEauthorrefmark{4}}
%\IEEEauthorblockA{\IEEEauthorrefmark{1}School of Electrical and Computer Engineering\\
%Georgia Institute of Technology,
%Atlanta, Georgia 30332--0250\\ Email: see http://www.michaelshell.org/contact.html}
%\IEEEauthorblockA{\IEEEauthorrefmark{2}Twentieth Century Fox, Springfield, USA\\
%Email: homer@thesimpsons.com}
%\IEEEauthorblockA{\IEEEauthorrefmark{3}Starfleet Academy, San Francisco, California 96678-2391\\
%Telephone: (800) 555--1212, Fax: (888) 555--1212}
%\IEEEauthorblockA{\IEEEauthorrefmark{4}Tyrell Inc., 123 Replicant Street, Los Angeles, California 90210--4321}}

% use for special paper notices
%\IEEEspecialpapernotice{(Invited Paper)}

% make the title area
\maketitle

\begin{abstract}

Rigorously validating a scientific model's explanatory power requires comparing its predictions against empirical data on an ongoing basis. 
However, model validation remains an informal and incomplete process, especially in fields like neuroscience where the available data is growing rapidly.   
This has made it difficult to develop a contemporary understanding of a model's strengths and weaknesses, to adequately compare competing models, to determine the state-of-the-art, and to precisely identify open modeling problems.

Software engineers validate software by writing simple executable tests, called ``unit tests'', that individually check a portion of a program against a single correctness criterion. 
Suites of such unit tests can collectively provide evidence for its validity and correctness. 
Drawing inspiration from this practice, we develop the \textit{SciUnit} framework for developing suites of ``model validation tests'' -- 
executable functions, here written in Python, that compare model predictions against a single empirical observation to produce a score indicating agreement between the model and the data. 
Suites of such validation tests are used as a summary of modeling progress within a scientific community. 
As a first application of this framework, we describe \textit{NeuronUnit}, a library that supports validation testing for neurophysiology and integrates with several existing neuroinformatics resources.

%in the context of the applications and architectures that the language may need to target.
%
%Programming languages targeted toward end-users in high-performance computing must consider a number of issues in addition to performance. Typically, researchers have focused verifiability, portability and ease-of-use, and the right balance often depends on the particular uses that they are put toward.
%
%Ace %, evaluates it against several design and adoption criteria, and describes several use cases as well as a realistic case study involving parallelizing neurobiological circuit simulations on clusters of GPUs. 
%is an extensible statically-typed programming language embedded within the ubiquitous scripting language Python. Python serves as a {metalanguage}, enabling programmatic control over several aspects of language specification and compilation. 
%We demonstrate the power of this design for HPC by defining all of the primitives of the OpenCL kernel language as a library of user-defined types in Ace, then extending this core with several higher-level first-class abstractions. The target of translation, called the \emph{backend}, is user-defined as well, allowing us to implement the semantics of these primitives by direct translation to OpenCL, CUDA or C99.
%
%Although Ace is built around a static type discipline, it supports a novel form of generic programming based on type propagation that, together with a form of extensible local type inference, eliminates most of the syntactic overhead usually associated with statically-typed languages. A companion host API allows type information from \verb|numpy| arrays to directly propagate into Ace kernels, supporting seamless adoption of Ace into current scientific workflows.
%
\end{abstract}

%\begin{IEEEkeywords}
%extensible programming languages, heterogeneous computing, type systems
%\end{IEEEkeywords}

% For peer review papers, you can put extra information on the cover
% page as needed:
% \ifCLASSOPTIONpeerreview
% \begin{center} \bfseries EDICS Category: 3-BBND \end{center}
% \fi
%
% For peerreview papers, this IEEEtran command inserts a page break and
% creates the second title. It will be ignored for other modes.
%\IEEEpeerreviewmaketitle

\section{Introduction}

Neuroscientists construct quantitative models to coherently explain observations of neurons, circuits, brain regions and behavior. 
These models can be characterized by their \textit{scope}: the set of observable quantities that the model can generate predictions about, and by their \textit{validity}: the extent to which these predictions agree with empirical observations of those quantities.

Today, scientists contribute a new model to the research community by submitting a paper describing how the model works, along with selected figures demonstrating its scope and validity and comparing its validity to other models with the same scope. 
Reviewers are then responsible for discovering relevant data and competing models that the paper did not adequately consider, drawing on their knowledge of prior publications. 
However, in many areas, the number of publications being generated every year can overwhelm even the most conscientious scientists \cite{jinha_article_2010}.

%This suggests that tools that help researchers answer questions like these could help readers and reviewers develop a more comprehensive understanding of the state-of-the-art and improve the quality of the research being published as a result: to renewed calls for tools that help researchers .

Unfortunately, there are few alternatives to a comprehensive literature review available when scientists need to answer questions like these:
\begin{itemize}
\item Which models are capable of predicting the quantities I am interested in?
\item Which metrics should be used to evaluate the goodness-of-fit between these models and  data?
\item How well do these models perform, as judged to these metrics, given currently available data?
\item What other quantities can and can't these models predict?
\item What observations are not adequately explained by any available model?
\end{itemize}

%These questions are particularly relevant for young scientists and  scientists aiming to enter a new research area to answer.

%understanding of the state-of-the-art from publications containing only a few pieces of data, and then only that data available at the time of publication.

%These demonstrations also quickly go out of date. Finally, there is no easy way to summarize and compare the claims being made in modeling papers in order to synthesize an understanding of a research area as a whole. 
%A strength of publications are their focused descriptions of new data and models. 
%A weakness, however, is that evaluating the scope and validity of models against known data is intractable using a body of publications alone.

Professional software developers face similar issues. 
They must understand the scope of each component of a complex software project and validate it by measuring how well each component achieves its specified input/output behavior. 
But software developers do not validate components by simply choosing a few interesting inputs and presenting the outputs to reviewers. 
Rather, they typically follow a \emph{test-driven development} methodology by creating a suite of executable \emph{unit tests} that serve to specify each component's scope and validate its implementation as it is being developed and modified \cite{beck2003}. 
Each test individually checks that a small portion of the program meets a single correctness criterion. For example, a unit test might verify that one function within the program correctly handles malformed inputs. 
Collectively, the test results serve as a summary of the validity of the project as it progresses through its development cycle. 
Developers can determine which features are unimplemented or buggy by examining the set of failed tests, and progress can be measured in terms of how many tests the program passes over time. 
This methodology is widely adopted in practice \cite{beck2003}.

%publications would be analagous to user manuals, rather than rigorous justifications that a program meets a specific specification. Unit tests serve this purpose for software, so we argue that the validation of neuroscience models by means of executable tests would be useful to scientists as a supplement the publication system.

Test-driven methodologies have started to see success in neuroscience as well. 
Modeling competitions in neuroscience, for example, are typically organized around a collection of simple validation criteria, implemented as executable tests. 
These competitions continue to drive important advances and improve scientists' understanding of the relative merits of different models. 
For example, the quantitative single neuron modeling competition (QSNMC) \cite{jolivet_quantitative_2008} investigates the complexity-accuracy tradeoff among reduced models of excitable membranes; 
the ``Hopfield'' challenge \cite{hopfield_what_2000} tested techniques for generating neuronal network form given its function; 
the Neural Prediction Challenge sought the best stimulus reconstructions, given neuronal activity (\url{http://neuralprediction.berkeley.edu}); the Diadem challenge is advancing the art of neurite reconstruction (\url{http://www.diademchallenge.org}); and 
examples from other subfields of biology abound (\url{http://www.the-dream-project.org}).
% \begin{figure}%{r}{0.6\textwidth}
% %\vspace{-38px}
% \includegraphics[scale=0.75]{sciunit_overview.pdf}
% \caption{Proposal overview. 
% \small{i) The \textit{SciUnit} framework can generate discipline-specific libraries for tests generation and model interfacing.  
% \textit{NeuronUnit} is proposed and described here.  Experimental data guides test development; model generation informs and is informed by these \textit{SciUnit} libraries.  
% Each \text{SciUnit} test suite repository on GitHub is indexed by the \textit{SciDash} web application.  
% Test suites for OSB and QSNMC (described in the text) are being built. 
% \textit{SciDash} automatically runs the test collections in each suite and publicly displays the results.}}
% %\vspace{-10px}
% \label{fig:sciunit_overview}
% \end{figure}
% \leavevmode
% \vspace{-0px}
\begin{figure*}[t]
\centering
\includegraphics[scale=0.4]{diagram1.pdf}
\vspace{-25px}
\caption{NeuronUnit overview. NeuronUnit is set of testing tools built upon the discipline-agnostic SciUnit framework. 
NeuronUnit can in principle test arbitrary neurophysiology models using arbitrary data but we provide here an example using models described in NeuroML as part of the \textit{Open Source Brain Project} (OSB, \cite{gleeson_open_2012}, \url{http://www.opensourcebrain.org}), and single neuron electrophysiology data available as part of the NeuroElectro Project (Neuroelectro, \cite{tripathy_neuroelectro:_2012}, \url{http://neuroelectro.org}).  
Records of test results for various model/test combinations are accessible via SciDash, which indexes GitHub repositories of these records, models, and tests so they can be searched and filtered by the community.}  
\label{fig:sciunit_overview}
\end{figure*}
\leavevmode
\vspace{-0px}

Each of these examples has leveraged \emph{ad hoc} infrastructure to support test generation. 
While the specific criteria used to evaluate models varies widely between disciplines in neuroscience, the underlying test-driven methodology has many common features that could be implemented once. 
Recognizing this, we developed a discipline-agnostic framework for developing scientific validation test suites called \textit{SciUnit} (\url{http://www.sciunit.org}). 
Here we describe \textit{NeuronUnit\-}, which builds upon \textit{SciUnit}, allowing neuroscientists to build \textit{SciUnit} tests that validate neurophysiology models against electrophysiological data.  We provide a concrete example pipeline, showing how models described using \text{NeuroML} and provided freely by the \textit{Open Source Brain Project} (OSB, \cite{gleeson_open_2012}, \url{http://www.opensourcebrain.org}) can be tested in fully automated fashion using published, curated data available through the \textit{NeuroElectro Project} (Neuroelectro, \cite{tripathy_neuroelectro:_2012}, \url{http://neuroelectro.org}), leveraging facilities from the \textit{NeuroTools} library (\url{http://neuralensemble.org/NeuroTools}) to extract relevant features of model output.  This is summarized in Figure \ref{fig:sciunit_overview}, which shows the relationships between the layers described here.
% In addition to this open testing toolkit we also provide a library of tests for models of spiking single neurons.

% Section 2.
\section{Validation Testing with \textit{SciUnit}}
\subsection{Example: The Quantitative Single Neuron Modeling Competition}
\begin{figure}
\small
\begin{python}
class SpikeCountTest(sciunit.Test):
  """Tests spike counts produced in response to several current stimuli against observed means and standard deviations.

  goodness of fit metric: Computes p-values based on a chi-squared test statistic, and pools them using Fisher's method.
  parameters:
    inputs: list of numpy arrays containing input currents (nA)
    means, stds: lists of observed means and standard deviations, one per input
  """
  def __init__(self, inputs, means, stds):
    self.inputs, self.means, self.stds = inputs, means, stds

  required_capabilities = [SpikeCountFromCurrent]

  def _judge(self, model):
    inputs, means, stds = self.inputs, self.means, self.stds
    n = len(inputs)
    counts = numpy.empty((n,))
    for i in xrange(n):
      counts[i] = model.spike_count_from_current(inputs[i])
    chisquared = sum((counts-means)**2 / means) # An array of chi-squared values.  
    p = scipy.stats.chi2.cdf(chisquared,n-1) # An array of p-values.  
    pooled_p = sciunit.utils.fisherp(p_array) # A pooled p-value.  
    return sciunit.PValue(pooled_p, related_data={
      "inputs": inputs, "counts": counts, "obs_means": means, "obs_stds": stds
    })
\end{python}
\vspace{-5px}
\caption{An example single neuron spike count test class implemented using \textit{SciUnit}. Because this implementation contains logic common to many different systems, \textit{NeuronUnit} was developed to provide a simpler means to deliver it (see Sec. \ref{sec:neuronunit}).}
\label{fig:rate_test}
\vspace{-15px}
\end{figure}

%Simple executable \emph{validation tests} that compute agreement between a model prediction and an experimental observation.  
We first illustrate the form of a generic example \textit{SciUnit} test suite that could be used in neurophysiology. 
Suppose we have collected data from an experiment where current stimuli (measured in pA) are delivered to neurons in some brain region, while the somatic membrane potential of each stimulated cell (in mV) is recorded and stored.  
A model claiming to capture this cell type's membrane potential dynamics must be able to accurately predict a variety of features observed in these data.

One simple validation test would ask candidate models to predict the number of action potentials (a.k.a. spikes) generated in response to a stimulus (e.g. white noise), and compare these \emph{spike count} predictions to the distribution observed in repeated experimental trials using the same stimulus. 
For data of this type, goodness-of-fit can be measured by first calculating a p-value from a chi-squared statistic for each prediction and then combining these p-values using Fisher's method \cite{fisher_statistical_1925}.

Alongside this \emph{spike count test}, we might also specify a number of other tests capturing different features of the data  to produce a more comprehensive suite. 
For data of this sort, the QSNMC defined 17 other validation criteria in addition to one based on the overall spike count, capturing features like spike latencies (SL), mean subthreshold voltage (SV), interspike intervals (ISI) and interspike minima (ISM) that can be extracted from the data \cite{jolivet_quantitative_2008}. 
They then defined a combined metric favoring models that broadly succeeded at meeting these criteria, to produce an overall ranking. 
Such combined criteria are simply validation tests that invoke other tests to produce a result.

%In many cases, models require no modifications to take the new tests because the same type of model output is being requested.
\subsection{Implementing a Validation Test in \textit{SciUnit}}
Fig. \ref{fig:rate_test} shows how a scientist can implement spike count tests such as the one described above using \textit{SciUnit}. 
A \textit{SciUnit} validation test is an {instance} (i.e. an object) of a Python class implementing the \verbx{sciunit.Test} interface (cf. line 1). 
Here, we show a class \verbx{SpikeCountTest} taking three \emph{parameters} in its constructor (constructors are named \verbx{__init__} in Python, lines 9-10). 
The meaning of each parameter along with a description of the goodness-of-fit metric used by the test is documented on lines 4-7. 
%For convenience, we also make use of functions provided by the popular NumPy\cite{numpy_url} and SciPy\cite{scipy_url} libraries, although these are not required by \textit{SciUnit}.  
To create a \emph{particular} spike count test, we instantiate this class with particular experimental observations. 
For example, given observations from hippocampal CA1 cells (not shown), we can instantiate a test as follows:
\begin{python}
  CA1_sc_test = SpikeCountTest(CA1_inputs, CA1_means, CA1_stds)
\end{python}
We emphasize the crucial distinction between the \textit{class} \verbx{SpikeCountTest}, which defines a \emph{parameterized family} of validation tests, and the particular \textit{instance} \verbx{CA1_sc_test}, which is an individual validation test because the necessary parameters, derived from data, have been provided. 
As we will describe below, we expect communities to build repositories of such families capturing the criteria used in their subfields of neuroscience so that test generation for a particular system of interest will often require simply instantiating a previously-developed family with particular experimental parameters and data. 
For single-neuron test families like \verbx{SpikeCountTest}, we have developed such a library, called \textit{NeuronUnit} (\url{http://github.com/scidash/neuronunit}) (Sec. \ref{sec:neuronunit}).

Classes that implement the \verbx{sciunit.Test} interface must contain a \verbx{_judge} method that receives a candidate \emph{model} as input and produces a \textit{score} as output. 
To specify the interface between the test and the model (that is, to specify an appropriate scope), the test author provides a list of \emph{capabilities} in the \verbx{required_capabilities} attribute, seen on line 12 of Fig. \ref{fig:rate_test}. 
Capabilities are simply collections of methods that a test will need to invoke in order to receive relevant data, and are analogous to \emph{interfaces} in e.g. Java (\url{http://docs.oracle.com/javase/tutorial/java/concepts/interface.html}). 
In Python, capabilities are written as classes with unimplemented members. 
The capability required by the test in Fig. \ref{fig:rate_test} is shown in Fig. \ref{fig:capability}. 
In \textit{SciUnit}, classes defining capabilities are tagged as such by inheriting from \verbx{sciunit.Capability}. The test in Figure \ref{fig:rate_test} uses this capability on line 19 to produce a spike count prediction for each input current.

The remainder of the \verbx{_judge} method implements the goodness-of-fit metric described above, returning an instance of \verbx{sciunit.PValue}, a subclass of \verbx{sciunit.Score} that is included with \textit{SciUnit}. In addition to the $p$-value itself, the returned score object also contains metadata, via the \verbx{related_data} parameter, for  scientists who may wish to examine the result in more detail later. 
In this case we save the input currents, the model outputs and the observed means and standard deviations (line 24). 
%We discuss visualization of results in Secs. \ref{sec:scidash_activities} and \ref{sec:scidash_visualization}.

\begin{figure}
\begin{python}
class SpikeCountFromCurrent(sciunit.Capability):
  def spike_count_from_current(self, input): 
    """Takes a numpy array containing current stimulus (in nA) and
    produces an integer spike count. Can be called multiple times."""
    raise NotImplementedError("Model does not implement capability.")
\end{python}
\caption{An example capability specifying a single required method (used by the test in Figure \ref{fig:rate_test}).}
\label{fig:capability}
\vspace{-10px}
\end{figure}

\begin{figure}
\begin{python}
class TrainSpikeCountFromCurrent(sciunit.Capability):
  def train_with_currents(self, currents, counts):
    """Takes a list of numpy arrays containing current stimulus (in nA) and
    observed spike counts. Model parameters should be adjusted based on this
    training data."""
    raise NotImplementedError("Model does not implement capability.")
\end{python}
\caption{Another capability specifying a training protocol (not used by the test in Figure \ref{fig:rate_test}).}
\label{fig:training}
\vspace{-15px}
\end{figure}

\subsection{Models}
Capabilities are \emph{implemented} by models. In \textit{SciUnit}, models are instances of Python classes that inherit from \verbx{sciunit.Model}. Like tests, the class itself represents a family of models, parameterized by the arguments of the constructor. A particular model is an instance of such a class.

Figure \ref{fig:simple_model} shows how to write a simple {family} of models, \verbx{LinearModel}, that implement the capability in Fig. \ref{fig:capability} as well as another capability shown in Fig. \ref{fig:training}, which we will discuss below. 
Models in this family generate a spike count by applying a linear transformation to the mean of the provided input current. The family is parameterized by the scale factor and the offset of the transformation, both scalars. 
To create a \emph{particular} linear model, a modeler can provide particular parameter values, just as with test families:
\begin{python}
CA1_linear_model_heuristic = LinearModel(3.0, 1.0)
\end{python}
Here, the parameters to the model were picked by the modeler heuristically, or based on externally-available knowledge. 
An alternative test design would add a training phase where these parameters were fit to data using the capability shown in Fig. \ref{fig:training}. 
This test could thus only be used for those models for which parameters can be adjusted without human involvement. 
Whether to build a training phase into the test protocol is a choice left to each test development community.

Fig. \ref{fig:rate_test} does not include a training phase. 
If training data is externally available, models that nevertheless do implement a training capability (like \verb|LinearModel|) can simply be trained explicitly by calling the capability method just like any other Python method:
\begin{python}
CA1_linear_model_fit = LinearModel()
CA1_linear_model_fit.train_with_currents(CA1_training_in, CA1_training_out)
\end{python}
\begin{figure}
\begin{python}
class LinearModel(sciunit.Model, SpikeCountFromCurrent, 
    TrainSpikeCountFromCurrent):
  def __init__(self, scale=None, offset=None): 
    self.scale, self.offset = scale, offset

  def spike_count_from_current(self, input):
    return int(self.scale*numpy.mean(input) + self.offset)

  def train_with_currents(self, currents, counts):
    means = [numpy.mean(c) for c in currents]
    [self.offset, self.scale] = numpy.polyfit(means, counts, deg=1)    
\end{python}
\caption{A model that returns a spike count by applying a linear transformation to the mean input current. The parameters can be provided manually or learned from data provided by a test or user (see text).}
\label{fig:simple_model}
\vspace{-15px}
\end{figure}

\subsection{Executing Tests} A test is executed against a model using the \verbx{judge} method:
\begin{python}
score = CA1_sc_test.judge(CA1_linear_model_heuristic)
\end{python}

This method proceeds by first checking that the provided model implements all required capabilities. It then calls the test's \verbx{_judge} method to produce a score. A reference to the test and model are added to the score for convenience (accessible via the \verbx{test} and \verbx{model} attributes, respectively), before it is returned.

\subsection{Test Suites and Score Matrices} A collection of tests intended to be run on the same model can be put together to form a test suite.
The following is a test suite that could be used for a simplified version of the QSNMC, as described above:  
\begin{python}
CA1_suite = sciunit.TestSuite([CA1_sc_test, CA1_sl_test, CA1_sv_test, CA1_isi_test, CA1_ism_test])
\end{python}
Like a single test, a test suite is capable of judging one or more models. The result is a score matrix much like the one diagramed in Fig. \ref{fig:sciunit_overview}.
\begin{python}
CA1_matrix = CA1_suite.judge([CA1_linear_model_heuristic, CA1_linear_model_fit])
\end{python}
A simple summary of the scores in a score matrix can be printed to the console or  visualized by other tools, such as the web application \textit{SciDash} described in Sec. \ref{sec:scidash}.

% conference papers do not normally have an appendix

% use section* for acknowledgement
% trigger a \newpage just before the given reference
% number - used to balance the columns on the last page
% adjust value as needed - may need to be readjusted if
% the document is modified later
%\IEEEtriggeratref{8}
% The "triggered" command can be changed if desired:
%\IEEEtriggercmd{\enlargethispage{-5in}}

% references section

% can use a bibliography generated by BibTeX as a .bbl file
% BibTeX documentation can be easily obtained at:
% http://www.ctan.org/tex-archive/biblio/bibtex/contrib/doc/
% The IEEEtran BibTeX style support page is at:
% http://www.michaelshell.org/tex/ieeetran/bibtex/
%\bibliographystyle{IEEEtran}
% argument is your BibTeX string definitions and bibliography database(s)
%\bibliography{IEEEabrv,../bib/paper}
%
% <OR> manually copy in the resultant .bbl file
% set second argument of \begin to the number of references
% (used to reserve space for the reference number labels box)
\bibliographystyle{IEEEtran}
\bibliography{../research}
%
%\begin{thebibliography}{1}
%
%\bibitem{IEEEhowto:kopka}
%H.~Kopka and P.~W. Daly, \emph{A Guide to \LaTeX}, 3rd~ed.\hskip 1em plus
%  0.5em minus 0.4em\relax Harlow, England: Addison-Wesley, 1999.
%
%\end{thebibliography}

% that's all folks
\end{document}

### 2013-11-22 sciunit-icse14/sciunit-icse14.tex
New abstract

@@
 \lstset{
-	basicstyle=\small\ttfamily,
+%	basicstyle=\tiny\ttfamily,
 	breaklines=true,
@@
 
-%\usepackage{alltt}
-%\renewcommand{\ttdefault}{txtt}
+\usepackage{alltt}
+\renewcommand{\ttdefault}{txtt}
 %%\usepackage{color}
@@
 \author{\alignauthor
-Cyrus Omar, Jonathan Aldrich\\\affaddr{Carnegie Mellon University, Pittsburgh, PA}\\
+Cyrus Omar, Jonathan Aldrich\\\affaddr{Carnegie Mellon University}\\
        \email{\{comar,aldrich\}@cs.cmu.edu}
-       \alignauthor Rick Gerkin\\
+       \alignauthor Richard C. Gerkin\\
        \affaddr{Arizona State University}\\{rgerkin@asu.edu}
@@
 \begin{abstract}
+One of the pillars of the modern scientific method is \emph{model validation}: comparing a scientific model's predictions against empirical observations.
+Today, a scientist demonstrates the validity of a model by making an argument in a paper and submitting it for peer review, a process comparable to \emph{code review} in software engineering.
+While peer review ensures that contributions meet high-level goals, software engineers typically supplement this process with \emph{unit testing} to get a more complete picture of the state of a software project, particularly for complex projects involving many developers.
 
-Rigorously validating a scientific model's explanatory power requires comparing its predictions against empirical data on an ongoing basis. 
-However, model validation remains an informal and incomplete process, especially in fields like neuroscience where the available data is growing rapidly.   
-This has made it difficult to develop a contemporary understanding of a model's strengths and weaknesses, to adequately compare competing models, to determine the state-of-the-art, and to precisely identify open modeling problems.
-
-Software engineers validate software by writing simple executable tests, called ``unit tests'', that individually check a portion of a program against a single correctness criterion. 
-Suites of such unit tests can collectively provide evidence for its validity and correctness. 
-Drawing inspiration from this practice, we develop the \textit{SciUnit} framework for developing suites of ``model validation tests'' -- 
-executable functions, here written in Python, that compare model predictions against a single empirical observation to produce a score indicating agreement between the model and the data. 
-Suites of such validation tests are used as a summary of modeling progress within a scientific community. 
-As a first application of this framework, we describe \textit{NeuronUnit}, a library that supports validation testing for neurophysiology and integrates with several existing neuroinformatics resources.
-
+We argue that a similar test-driven methodology would be valuable to scientific communities that need to validate increasingly complex models against growing collections of empirical data.
+The dynamics of scientific communities differ from software communities in several key ways, however.
+In this paper, we introduce \emph{SciUnit}, a test-driven model validation framework designed to handle these differences in a practical manner.
+We also describe how SciUnit, supported by both new and existing collaborative infrastructure, can be integrated into the modern scientific process.
 %in the context of the applications and architectures that the language may need to target.
@@
 \subsection{Example: The Quantitative Single Neuron Modeling Competition}
-\begin{figure}
+\begin{figure}[t]
 \small

### 2013-11-23 sciunit-icse14/sciunit-icse14.tex
First few paragraphs of new intro

@@
 Today, a scientist demonstrates the validity of a model by making an argument in a paper and submitting it for peer review, a process comparable to \emph{code review} in software engineering.
-While peer review ensures that contributions meet high-level goals, software engineers typically supplement this process with \emph{unit testing} to get a more complete picture of the state of a software project, particularly for complex projects involving many developers.
+While human review helps to ensure that contributions meet high-level goals, software engineers typically supplement this with \emph{unit testing} to get a more complete picture of the status of a software project, particularly for complex projects involving many developers.
 
-We argue that a similar test-driven methodology would be valuable to scientific communities that need to validate increasingly complex models against growing collections of empirical data.
-The dynamics of scientific communities differ from software communities in several key ways, however.
-In this paper, we introduce \emph{SciUnit}, a test-driven model validation framework designed to handle these differences in a practical manner.
-We also describe how SciUnit, supported by both new and existing collaborative infrastructure, can be integrated into the modern scientific process.
+We argue that a similar test-driven methodology would be valuable to scientific communities as they seek to  validate increasingly complex models against growing collections of empirical data.
+The dynamics of  scientific communities and software communities differ in several key ways, however.
+In this paper, we introduce \emph{SciUnit}, a framework for test-driven scientific model validation.
+We describe how SciUnit, supported by new and existing collaborative infrastructure, can integrate into the modern scientific process.
 %in the context of the applications and architectures that the language may need to target.
@@
 \section{Introduction}
+Scientific theories often take the form of a \emph{quantitative model}: a formal structure that can generate predictions about observable quantities.
+Such a model is characterized by its \textit{scope}: the set of observable quantities that the model can predict, and by its \textit{validity}: the extent to which its predictions agree with experimental observations of these quantities.
 
-Neuroscientists construct quantitative models to coherently explain observations of neurons, circuits, brain regions and behavior. 
-These models can be characterized by their \textit{scope}: the set of observable quantities that the model can generate predictions about, and by their \textit{validity}: the extent to which these predictions agree with empirical observations of those quantities.
+Quantitative models are described and validated using a social process: the scientific publication system.
+For a model to be accepted by the scientific community, scientists must write a paper describing it and providing  evidence that it predicts some quantity of interest more accurately than previous models, or that the tradeoff it makes between accuracy and complexity may be useful to the community \cite{box}.
+During the \emph{peer review} process, other members of the relevant community are tasked with ensuring that the validation criteria used (e.g.
+a $p$-value) are strong enough and that all relevant data and competing models were fully considered, drawing on knowledge of statistical methods and the prior literature.
+Publishing a paper is one of the primary motivators for most scientists \cite{howison2011scientific}.
 
-Today, scientists contribute a new model to the research community by submitting a paper describing how the model works, along with selected figures demonstrating its scope and validity and comparing its validity to other models with the same scope.
-Reviewers are then responsible for discovering relevant data and competing models that the paper did not adequately consider, drawing on their knowledge of prior publications. 
+Quantitative models have much in common with software.
+Indeed, quantitative models are increasingly  implemented as computer programs.
+In some cases the program implements a mathematical model, while in other cases the program \emph{is} the model (e.g.
+complex simulations).
+The peer review process is similar in many ways to the \emph{code review} process used by many development teams \cite{codereview}.
+During code review, team members look for errors, enforce style and architectural guidelines and check that the code is valid (i.e.
+that achieves its intended goal \cite{validity}) before it is committed to a primary source code repository.
+Code review is reasonably effective \cite{creffectiveness1,creffectiveness2}.
+However, to be most effective reviewers must expend considerable effort \cite{effort}, and code review is most effective for resolving  issues related to software architecture \cite{evolvability}.
+
+For these reasons, software development teams typically supplement code reviews with a number of more \emph{automated} approaches. One of the most widely-adopted approaches is \emph{unit testing}.
+
+--------
+
 However, in many areas, the number of publications being generated every year can overwhelm even the most conscientious scientists \cite{jinha_article_2010}.

### 2013-11-23 sciunit-icse14/sciunit-icse14.tex
First few paragraphs of new intro

@@
 
-Quantitative models have much in common with software.
+Quantitative scientific modeling and software development have much in common.
 Indeed, quantitative models are increasingly implemented as computer programs.
-In some cases the program implements a mathematical model, while in other cases the program \emph{is} the model (e.g.
+In some cases, the program \emph{is} the model (e.g.
 complex simulations).
-The peer review process is similar in many ways to the \emph{code review} process used by many development teams \cite{codereview}.
-During code review, team members look for errors, enforce style and architectural guidelines and check that the code is valid (i.e.
-that achieves its intended goal \cite{validity}) before it is committed to a primary source code repository.
+The peer review process is similar in many ways to the \emph{code review} process used in many development teams.
+During code review, team members look for errors, enforce style and architectural guidelines and check that the code is \emph{valid} (i.e.
+that achieves its intended goal \cite{validity}) before it can be committed to the primary source code repository \cite{codereview}.
+
 Code review is reasonably effective \cite{creffectiveness1,creffectiveness2}.
 However, to be most effective reviewers must expend considerable effort \cite{effort}, and code review is most effective for resolving  issues related to software architecture \cite{evolvability}.
-
-For these reasons, software development teams typically supplement code reviews with a number of more \emph{automated} approaches. One of the most widely-adopted approaches is \emph{unit testing}.
+For these reasons, most software development teams supplement code reviews with a number of more {automated} approaches to verification and validation.
+One of the most widely-adopted approaches is \emph{unit testing}.
 

### 2013-11-23 sciunit-icse14/sciunit-icse14.tex
Started describing unit testing

@@
 One of the most widely-adopted approaches is \emph{unit testing}.
+In brief, unit tests are functions that check a single component of a program against a single well-defined correctness criterion.
+A suite of such tests can be seen as a partial specification of a program or component.
+Test suites complement code review procedures in several ways:
+\begin{enumerate}
+\item They help ensure that a code modification does not cause a \emph{regression} in another part of a program.
+\item They help measure development progress. Developers can use a test suite to determine which functionality is not yet adequately implemented.
+\item They serve as a form of documentation. By inspecting the test suite itself, a developer can see which modes of usage the development team collectively considers most important.
+\end{enumerate}
 

### 2013-11-23 sciunit-icse14/sciunit-icse14.tex
Started describing unit testing

--- 2013-11-23 sciunit-icse14/sciunit-icse14.tex
@@
 One of the most widely-adopted approaches is \emph{unit testing}.
+In brief, unit tests are functions that check a single component of a program against a single well-defined correctness criterion \cite{beck2003}.
+A suite of such tests can be seen as a partial specification of a program or component.
+Test suites complement code review by allowing developers to answer questions like these more easily:
+\begin{enumerate}
+\item How much development progress has been made? Which functionality has not yet been adequately implemented? How valuable is a code contribution?
+\item What modes of usage does the team consider most important? How does the development team measure success?
+\item Does a code contribution cause \emph{regressions} in other parts of a program?
+\end{enumerate}
+Scientists ask analagous questions:
+\begin{enumerate}
+\item When new experimental observations are made, scientists want to determine the impact they have on the validity of previously-developed models.
+\end{enumerate}
+Answering these questions is often difficult using scientific publications alone.
+Each publication is frozen in time, so its examination of validity cannot consider the most recent body of data, or modern validation criteria.
+Discovering these validation criteria, as well as precisely characterizing open modeling problems  requires an encyclopedia knowledge of the literature.
+In many areas, the number of publications generated every year can overwhelm even the most conscientious and experienced scientists \cite{jinha_article_2010}.
+A common complaint of statisticians is that scientists are not aware of best practices and that community standards evolve too slowly \cite{grumpystatisticians}.
 
---------
-
-However, in many areas, the number of publications being generated every year can overwhelm even the most conscientious scientists \cite{jinha_article_2010}.
+Unfortunately, there are few alternatives to comprehensive literature review available when scientists need to answer questions like these today.%Anecdotally, one of the greatest fears of new scientists is that a problem that they are working on has already been solved and published.
 
@@
 
-Unfortunately, there are few alternatives to a comprehensive literature review available when scientists need to answer questions like these:
+-------
+
 \begin{itemize}
@@
 %A weakness, however, is that evaluating the scope and validity of models against known data is intractable using a body of publications alone.
-
-Professional software developers face similar issues. 
-They must understand the scope of each component of a complex software project and validate it by measuring how well each component achieves its specified input/output behavior. 
-But software developers do not validate components by simply choosing a few interesting inputs and presenting the outputs to reviewers. 
-Rather, they typically follow a \emph{test-driven development} methodology by creating a suite of executable \emph{unit tests} that serve to specify each component's scope and validate its implementation as it is being developed and modified \cite{beck2003}.
-Each test individually checks that a small portion of the program meets a single correctness criterion.
-For example, a unit test might verify that one function within the program correctly handles malformed inputs.
-Collectively, the test results serve as a summary of the validity of the project as it progresses through its development cycle. 
-Developers can determine which features are unimplemented or buggy by examining the set of failed tests, and progress can be measured in terms of how many tests the program passes over time. 
-This methodology is widely adopted in practice \cite{beck2003}.
 

### 2013-11-23 sciunit-icse14/sciunit-icse14.tex
Started describing unit testing
more progress on intro

@@
 
-Quantitative models are described and validated using a social process: the scientific publication system.
-For a model to be accepted by the scientific community, scientists must write a paper describing it and providing  evidence that it predicts some quantity of interest more accurately than previous models, or that the tradeoff it makes between accuracy and complexity may be useful to the community \cite{box}.
-During the \emph{peer review} process, other members of the relevant community are tasked with ensuring that the validation criteria used (e.g.
-a $p$-value) are strong enough and that all relevant data and competing models were fully considered, drawing on knowledge of statistical methods and the prior literature.
-Publishing a paper is one of the primary motivators for most scientists \cite{howison2011scientific}.
+Quantitative models are validated by \emph{peer review}.
+For a model to be accepted by the scientific community, scientists must submit a paper describing it and providing  evidence that it predicts some quantity of interest more accurately than previous models, or that it makes a desirable tradeoff  between accuracy and complexity \cite{box}.
+Other members of the relevant community are then tasked with ensuring that validity was measured properly and that relevant data and competing models were adequately considered, drawing on knowledge of statistical methods and the prior literature.
+Publishing is a primary motivator for most scientists \cite{howison2011scientific}.
 
 Quantitative scientific modeling and software development have much in common.
-Indeed, quantitative models are increasingly implemented as computer programs.
-In some cases, the program \emph{is} the model (e.g.
+Indeed, quantitative models are increasingly  implemented as computer programs and in some cases, the program \emph{is} the model (e.g.
 complex simulations).
-The peer review process is similar in many ways to the \emph{code review} process used in many development teams.
-During code review, team members look for errors, enforce style and architectural guidelines and check that the code is \emph{valid} (i.e.
-that achieves its intended goal \cite{validity}) before it can be committed to the primary source code repository \cite{codereview}.
+The peer review process for papers is similar in many ways to the \emph{code review} process used in many development teams, where team members look for errors, enforce style and architectural guidelines and check that the code is \emph{valid} (i.e.
+that achieves its intended goal \cite{validity}) before permitting it to be committed to the primary source code repository \cite{codereview}.
 
-Code review is reasonably effective \cite{creffectiveness1,creffectiveness2}.
-However, to be most effective reviewers must expend considerable effort \cite{effort}, and code review is most effective for resolving  issues related to software architecture \cite{evolvability}.
-For these reasons, most software development teams supplement code reviews with a number of more {automated} approaches to verification and validation.
-One of the most widely-adopted approaches is \emph{unit testing}.
-In brief, unit tests are functions that check a single component of a program against a single well-defined correctness criterion \cite{beck2003}.
+Code review can be quite effective \cite{creffectiveness1,creffectiveness2}, but this requires that developers  expend considerable effort \cite{effort}.
+Code review is also most effective for resolving  issues related to software architecture \cite{evolvability}.
+Most development teams thus supplement code reviews with {automated} approaches to verification and validation, the most widely-used  of which is \emph{unit testing} \cite{beck2003}.
+In brief, unit tests are functions that check a single portion of a program against a single well-defined correctness criterion.
 A suite of such tests can be seen as a partial specification of a program or component.
@@
 \begin{enumerate}
-\item How much development progress has been made? Which functionality has not yet been adequately implemented? How valuable is a code contribution?
-\item What modes of usage does the team consider most important? How does the development team measure success?
+\item Which functionality has been adequately implemented? What remains to be done?
+%\item How much progress does an individual code contribution make? 
+\item What modes of usage does the team consider most important? How does the team measure correctness?
 \item Does a code contribution cause \emph{regressions} in other parts of a program?
@@
 \begin{enumerate}
-\item When new experimental observations are made, scientists want to determine the impact they have on the validity of previously-developed models.
+\item Which observations are already well-explained by existing models? What is the state-of-the-art? What are the open modeling problems of interest?
+%\item How much better is a new model than the state of the art? What are the open problems in my field?
+\item What are the contemporary community standards for measuring goodness-of-fit?
+\item How do newly-made experimental observations impact the validity of previous models?
 \end{enumerate}
-Answering these questions is often difficult using scientific publications alone.
-Each publication is frozen in time, so its examination of validity cannot consider the most recent body of data, or modern validation criteria.
-Discovering these validation criteria, as well as precisely characterizing open modeling problems  requires an encyclopedia knowledge of the literature.
-In many areas, the number of publications generated every year can overwhelm even the most conscientious and experienced scientists \cite{jinha_article_2010}.
-A common complaint of statisticians is that scientists are not aware of best practices and that community standards evolve too slowly \cite{grumpystatisticians}.
+But while software engineers can rely on a program's test suite, scientists today must extract this information from a body of scientific publications.
+This is increasingly difficult.
+Each publication focuses on just one model and is frozen in time, so it does not consider the latest experimental data or statistical methods.
+Discovering, precisely characterizing and comparing several models to discover the state of the art, and finding open modeling problems, often requires an encyclopedic knowledge of the literature.
+Senior scientists can attempt to summarize this with review papers, but in many areas, the number of publications generated every year can be overwhelming \cite{jinha_article_2010}, and comprehensive reviews are published relatively infrequently for any given area.
+Statisticians commonly complain that scientists are not following best practices and that community standards evolve too slowly because canonical papers or popular reviews advocate outdated methods \cite{grumpystatisticians}.
+One might compare this to a  software development team relying primarily on carefully-prepared minutes of code review sessions, peppered with cross-references to previous minutes, to understand and validate a complex software project.
 
-Unfortunately, there are few alternatives to comprehensive literature review available when scientists need to answer questions like these today.%Anecdotally, one of the greatest fears of new scientists is that a problem that they are working on has already been solved and published.
+We suggest that scientific model validation could benefit from the adoption of a test-driven methodology.
+However, scientific communities differ from typical development teams in several important ways.
+
+- Unit tests are typically pass/fail, while goodness-of-fit is typically measured more granularly. 
+- Different communities use different goodness-of-fit measures. These continuously evolve as, for example, statisticians develop better methods.
+- Professional software developers are used to testing, while scientists are professional end-users and rarely use testing, even for verification.
+
+%Anecdotally, one of the greatest fears of new scientists is that a problem that they are working on has already been solved and published.
 
 %This suggests that tools that help researchers answer questions like these could help readers and reviewers develop a more comprehensive understanding of the state-of-the-art and improve the quality of the research being published as a result: to renewed calls for tools that help researchers .
-
--------
-
-\begin{itemize}
-\item Which models are capable of predicting the quantities I am interested in?
-\item Which metrics should be used to evaluate the goodness-of-fit between these models and  data?
-\item How well do these models perform, as judged to these metrics, given currently available data?
-\item What other quantities can and can't these models predict?
-\item What observations are not adequately explained by any available model?
-\end{itemize}
 

### 2013-11-23 sciunit-icse14/sciunit-icse14.tex
more progress on intro

@@
 \section{Introduction}
+\begin{figure*}[t]
+\centering
+\includegraphics[scale=0.4]{diagram1.pdf}
+\vspace{-25px}
+\caption{NeuronUnit overview. NeuronUnit is set of testing tools built upon the discipline-agnostic SciUnit framework. 
+NeuronUnit can in principle test arbitrary neurophysiology models using arbitrary data but we provide here an example using models described in NeuroML as part of the \textit{Open Source Brain Project} (OSB, \cite{gleeson_open_2012}, \url{http://www.opensourcebrain.org}), and single neuron electrophysiology data available as part of the NeuroElectro Project (Neuroelectro, \cite{tripathy_neuroelectro:_2012}, \url{http://neuroelectro.org}).
+Records of test results for various model/test combinations are accessible via SciDash, which indexes GitHub repositories of these records, models, and tests so they can be searched and filtered by the community.}
+\label{fig:sciunit_overview}
+\end{figure*}
+
 Scientific theories often take the form of a \emph{quantitative model}: a formal structure that can generate predictions about observable quantities.
@@
 Each publication focuses on just one model and is frozen in time, so it does not consider the latest experimental data or statistical methods.
-Discovering, precisely characterizing and comparing several models to discover the state of the art, and finding open modeling problems, often requires an encyclopedic knowledge of the literature.
-Senior scientists can attempt to summarize this with review papers, but in many areas, the number of publications generated every year can be overwhelming \cite{jinha_article_2010}, and comprehensive reviews are published relatively infrequently for any given area.
+Discovering, precisely characterizing and comparing several models to discover the state of the art and find open modeling problems often requires an encyclopedic knowledge of the literature.
+Senior scientists can attempt to summarize this by writing review papers, but in many areas, the number of publications generated every year can be overwhelming \cite{jinha_article_2010}, and comprehensive reviews are published relatively infrequently for any given area.
 Statisticians commonly complain that scientists are not following best practices and that community standards evolve too slowly because canonical papers or popular reviews advocate outdated methods \cite{grumpystatisticians}.
-One might compare this to a  software development team relying primarily on carefully-prepared minutes of code review sessions, peppered with cross-references to previous minutes, to understand and validate a complex software project.
+One might compare this process to a software development team relying primarily on carefully-prepared summaries of code review sessions, peppered with cross-references to previous summaries, to understand and validate a complex software project.
 
-We suggest that scientific model validation could benefit from the adoption of a test-driven methodology.
-However, scientific communities differ from typical development teams in several important ways.
-
-- Unit tests are typically pass/fail, while goodness-of-fit is typically measured more granularly. 
-- Different communities use different goodness-of-fit measures. These continuously evolve as, for example, statisticians develop better methods.
-- Professional software developers are used to testing, while scientists are professional end-users and rarely use testing, even for verification.
-
+This suggests that quantitative model validation could be improved by the adoption of test-driven methodologies alongside peer review.
+However, the scientific community presents several unique challenges:
+\begin{enumerate}
+\item Unit tests are typically pass/fail, while goodness-of-fit is typically a continuous value (e.g. a $p$-value). 
+\item Different communities prefer different goodness-of-fit measures. These continuously evolve as, for example, statisticians develop better methods.
+\item Model developers compete with one another to predict data, rather than collaborating with one another on a common codebase.
+They are sometimes reluctant to share certain details of their implementations \cite{badscientists}.
+\item Experimental data can itself be erroneous, noisy or inconsistent with existing data.
+\item Data formats differ, sometimes drastically, between experimentalists. Similarly, different modelers may prefer different programming languages or libraries.
+\item Professional software developers are typically trained in testing practices and tools, while scientists  rarely use testing practices with their own code \cite{badscientists2}.
+\end{enumerate}
 %Anecdotally, one of the greatest fears of new scientists is that a problem that they are working on has already been solved and published.
@@
 Unit tests serve this purpose for software, so we argue that the validation of neuroscience models by means of executable tests would be useful to scientists as a supplement the publication system.
-
 Test-driven methodologies have started to see success in neuroscience as well. 
@@
 % \vspace{-0px}
-\begin{figure*}[t]
-\centering
-\includegraphics[scale=0.4]{diagram1.pdf}
-\vspace{-25px}
-\caption{NeuronUnit overview. NeuronUnit is set of testing tools built upon the discipline-agnostic SciUnit framework. 
-NeuronUnit can in principle test arbitrary neurophysiology models using arbitrary data but we provide here an example using models described in NeuroML as part of the \textit{Open Source Brain Project} (OSB, \cite{gleeson_open_2012}, \url{http://www.opensourcebrain.org}), and single neuron electrophysiology data available as part of the NeuroElectro Project (Neuroelectro, \cite{tripathy_neuroelectro:_2012}, \url{http://neuroelectro.org}).
-Records of test results for various model/test combinations are accessible via SciDash, which indexes GitHub repositories of these records, models, and tests so they can be searched and filtered by the community.}
-\label{fig:sciunit_overview}
-\end{figure*}
-\leavevmode
-\vspace{-0px}
 

### 2013-11-23 sciunit-icse14/sciunit-icse14.tex
more progress on intro

@@
 Each publication focuses on just one model and is frozen in time, so it does not consider the latest experimental data or statistical methods.
-Discovering, precisely characterizing and comparing several models to discover the state of the art and find open modeling problems often requires an encyclopedic knowledge of the literature.
-Senior scientists can attempt to summarize this by writing review papers, but in many areas, the number of publications generated every year can be overwhelming \cite{jinha_article_2010}, and comprehensive reviews are published relatively infrequently for any given area.
+Discovering, precisely characterizing and comparing several models to discover the state of the art and find open modeling problems can often require an encyclopedic knowledge of the literature.
+Senior scientists can attempt to fill this need by publishing review papers, but in many areas, the number of publications generated every year can be overwhelming \cite{jinha_article_2010}, and comprehensive reviews are published relatively infrequently for any given area.
 Statisticians commonly complain that scientists are not following best practices and that community standards evolve too slowly because canonical papers or popular reviews advocate outdated methods \cite{grumpystatisticians}.
@@
 
-This suggests that quantitative model validation could be improved by the adoption of test-driven methodologies alongside peer review.
-However, the scientific community presents several unique challenges:
+This suggests that the scientific process could be improved by the adoption of test-driven methodologies alongside peer review.
+However, the scientific community presents several unique challenges that must first be addressed:
 \begin{enumerate}
 \item Unit tests are typically pass/fail, while goodness-of-fit is typically a continuous value (e.g. a $p$-value). 
-\item Different communities prefer different goodness-of-fit measures. These continuously evolve as, for example, statisticians develop better methods.
+\item Different communities, groups and individual scientists prefer different goodness-of-fit measures and focus on different sets of observable quantities.
+This can evolve as, for example, statisticians develop better measures.
 \item Model developers compete with one another to predict data, rather than collaborating with one another on a common codebase.
 They are sometimes reluctant to share certain details of their implementations \cite{badscientists}.
-\item Experimental data can itself be erroneous, noisy or inconsistent with existing data.
-\item Data formats differ, sometimes drastically, between experimentalists. Similarly, different modelers may prefer different programming languages or libraries.
-\item Professional software developers are typically trained in testing practices and tools, while scientists  rarely use testing practices with their own code \cite{badscientists2}.
+\item Experimental data can itself be erroneous, noisy or inconsistent with other available data.
+\item Data formats can differ, sometimes drastically, between experimentalists. Similarly, different modelers may prefer different programming languages or libraries.
+\item Professional software developers are typically trained in testing practices and tools, while scientists  rarely have training or experience with testing practices \cite{badscientists2}.
 \end{enumerate}

### 2013-11-23 sciunit-icse14/sciunit-icse14.tex
updated figure caption

@@
 \vspace{-25px}
-\caption{NeuronUnit overview. NeuronUnit is set of testing tools built upon the discipline-agnostic SciUnit framework. 
-NeuronUnit can in principle test arbitrary neurophysiology models using arbitrary data but we provide here an example using models described in NeuroML as part of the \textit{Open Source Brain Project} (OSB, \cite{gleeson_open_2012}, \url{http://www.opensourcebrain.org}), and single neuron electrophysiology data available as part of the NeuroElectro Project (Neuroelectro, \cite{tripathy_neuroelectro:_2012}, \url{http://neuroelectro.org}).
-Records of test results for various model/test combinations are accessible via SciDash, which indexes GitHub repositories of these records, models, and tests so they can be searched and filtered by the community.}
+\caption{The central table summarizes the performance of a collection of models against a suite of tests.
+These are collaboratively curated in \emph{suite repositories} using social coding infrastructure -- Github -- combined with a portal that organizes and summarizes them.
+Tests are derived from data and models are derived from scientific theories, parameterized by \emph{training data}.
+Tests communicate with models using interface specifications that are also collaboratively developed in repositories that build upon the core SciUnit testing framework.
+These repositories can also contain common test and model families and bridges to existing infrastructure and tools.}
 \label{fig:sciunit_overview}

### 2013-11-23 sciunit-icse14/sciunit-icse14.tex
Substantial revision of figure 1
Final NIER submission

--- 2013-11-23 sciunit-icse14/sciunit-icse14.tex
@@
 \lstset{
-%	basicstyle=\tiny\ttfamily,
+	basicstyle=\footnotesize\ttfamily,
 	breaklines=true,
@@
 	}
-\let\verbx\lstinline
+\newcommand{\verbx}[1]{\lstinline{#1}}
 \newcommand{\bfhead}[1]{\noindent \textbf{#1:}}
@@
 % can use linebreaks \\ within to get better formatting as desired
-\title{Collaborative Test-Driven Scientific Model Validation}
+\title{Collaborative Infrastructure for Test-Driven\\Scientific Model Validation}
 
@@
 %\begin{IEEEkeywords}
-%extensible programming languages, heterogeneous computing, type systems
+%unit testing, validation, computer-supported collaborative work
 %\end{IEEEkeywords}
@@
 \begin{figure*}[t]
+\vspace{-30px}
 \centering
-\includegraphics[scale=0.4]{diagram1.pdf}
-\vspace{-25px}
-\caption{NeuronUnit overview. NeuronUnit is set of testing tools built upon the discipline-agnostic SciUnit framework. 
-NeuronUnit can in principle test arbitrary neurophysiology models using arbitrary data but we provide here an example using models described in NeuroML as part of the \textit{Open Source Brain Project} (OSB, \cite{gleeson_open_2012}, \url{http://www.opensourcebrain.org}), and single neuron electrophysiology data available as part of the NeuroElectro Project (Neuroelectro, \cite{tripathy_neuroelectro:_2012}, \url{http://neuroelectro.org}).
-Records of test results for various model/test combinations are accessible via SciDash, which indexes GitHub repositories of these records, models, and tests so they can be searched and filtered by the community.}
+\includegraphics[scale=0.45]{diagram1.pdf}
+\vspace{-40px}
+\caption{Tests are derived from data and models are derived from scientific theories.
+The score table summarizes the performance of a collection of models against a suite of tests.
+A table is contained in a \emph{suite repository} (e.g.
+\texttt{saturnsuite}), hosted on social coding infrastructure -- here, Github.
+SciDash is a portal that discovers and organizes these suite repositories  and provides tools for visualizing them.
+Tests interface with models using capabilities, as specified by the testing framework (\texttt{sciunit}).
+Common testing code, capabilities and bridges to external model and data repositories are also collaboratively developed (e.g.
+\texttt{cosmounit}).}
 \label{fig:sciunit_overview}
@@
 Quantitative models are validated by \emph{peer review}.
-For a model to be accepted by the scientific community, scientists must submit a paper describing it and providing  evidence that it predicts some quantity of interest more accurately than previous models, or that it makes a desirable tradeoff  between accuracy and complexity \cite{box}.
+For a model to be accepted by the scientific community, scientists must submit a paper describing it and providing  evidence that it predicts some quantity of interest more accurately than previous models, or that it makes a desirable tradeoff  between accuracy and complexity \cite{box1987empirical}.
 Other members of the relevant community are then tasked with ensuring that validity was measured properly and that relevant data and competing models were adequately considered, drawing on knowledge of statistical methods and the prior literature.
@@
 The peer review process for papers is similar in many ways to the \emph{code review} process used in many development teams, where team members look for errors, enforce style and architectural guidelines and check that the code is \emph{valid} (i.e.
-that achieves its intended goal \cite{validity}) before permitting it to be committed to the primary source code repository \cite{codereview}.
+that it achieves its intended goal) before permitting it to be committed to the primary source code repository.
 
-Code review can be quite effective \cite{creffectiveness1,creffectiveness2}, but this requires that developers  expend considerable effort \cite{effort}.
-Code review is also most effective for resolving  issues related to software architecture \cite{evolvability}.
+Code review can be quite effective \cite{codereview}, but this requires that developers  expend considerable effort \cite{kemerer2009impact}.
+Code review is also most effective for resolving  issues related to software architecture.
 Most development teams thus supplement code reviews with {automated} approaches to verification and validation, the most widely-used  of which is \emph{unit testing} \cite{beck2003}.
@@
 \item What are the contemporary community standards for measuring goodness-of-fit?
-\item How do newly-made experimental observations impact the validity of previous models?
+\item How do newly-made experimental observations impact the validity of previously-published models?
 \end{enumerate}
@@
 Each publication focuses on just one model and is frozen in time, so it does not consider the latest experimental data or statistical methods.
-Discovering, precisely characterizing and comparing several models to discover the state of the art and find open modeling problems often requires an encyclopedic knowledge of the literature.
-Senior scientists can attempt to summarize this by writing review papers, but in many areas, the number of publications generated every year can be overwhelming \cite{jinha_article_2010}, and comprehensive reviews are published relatively infrequently for any given area.
-Statisticians commonly complain that scientists are not following best practices and that community standards evolve too slowly because canonical papers or popular reviews advocate outdated methods \cite{grumpystatisticians}.
-One might compare this process to a software development team relying primarily on carefully-prepared summaries of code review sessions, peppered with cross-references to previous summaries, to understand and validate a complex software project.
+Discovering, precisely characterizing and comparing models to discover the state of the art and find open modeling problems  require an encyclopedic knowledge of the literature.
+Senior scientists can attempt to fill this need by publishing review papers, but in many areas, the number of publications generated every year can be overwhelming \cite{jinha_article_2010}, and comprehensive reviews of a particular area are published relatively infrequently.
+Statisticians often complain that scientists are not following best practices and that community standards evolve too slowly because a canonical paper or popular review used outdated methods.
+Furthermore, if the literature simply doesn't address an important question of validity, a researcher often needs to reimplement a model from scratch.
 
-This suggests that quantitative model validation could be improved by the adoption of test-driven methodologies alongside peer review.
-However, the scientific community presents several unique challenges:
+One might compare this to a software development team relying primarily on carefully-prepared and cross-referenced summaries of code review sessions to understand, validate and evolve a complex software project where  portions of the original source code have been obscured or lost.
+Although certainly a caricature, this motivates our suggestion that the scientific process could be improved by the adoption of test-driven methodologies alongside traditional peer review.
+However, the scientific community presents several challenges that must be addressed by any such methodology:
 \begin{enumerate}
-\item Unit tests are typically pass/fail, while goodness-of-fit is typically a continuous value (e.g. a $p$-value). 
-\item Different communities prefer different goodness-of-fit measures. These continuously evolve as, for example, statisticians develop better methods.
-\item Model developers compete with one another to predict data, rather than collaborating with one another on a common codebase.
-They are sometimes reluctant to share certain details of their implementations \cite{badscientists}.
-\item Experimental data can itself be erroneous, noisy or inconsistent with existing data.
-\item Data formats differ, sometimes drastically, between experimentalists. Similarly, different modelers may prefer different programming languages or libraries.
-\item Professional software developers are typically trained in testing practices and tools, while scientists  rarely use testing practices with their own code \cite{badscientists2}.
+\item Unit tests are typically pass/fail, while goodness-of-fit between a model and data is typically measured by a continuous metric (e.g. a $p$-value). 
+\item Unit tests often test a \emph{particular} component, whereas a \emph{validation test} must be able to uniformly handle any model capable of predicting the quantity being tested.
+%They are sometimes reluctant to share certain details of their implementations \cite{badscientists}.
+%\item Experimental data can itself be erroneous, noisy or inconsistent with other available data.
+\item Different modelers often  prefer different programming languages or frameworks.
+Similarly, data formats can differ between experimentalists.
+The testing framework must be as flexible as possible about these choices.
+\item Professional software developers are typically trained in testing practices and tools, while scientists  rarely have training or experience with testing practices \cite{oai:open.ac.uk.OAI2:17673}.
+Thus, the framework must be as simple as possible.
+\item Different communities, groups and individuals prefer different goodness-of-fit metrics and emphasize different sets of observable quantities.
+These can evolve as, for example, statisticians develop better measures.
+In contrast, there is more pressure to agree upon requirements and priorities in a software development project.
 \end{enumerate}
+
+To address challenges 1-4, we will introduce a lightweight scientific validation testing framework, \textit{SciUnit}, in Sec.
+2.
+Challenge 5 has to do with coordination between scientists.
+To address this, we introduce a community workflow based on widely-adopted social coding tools (here Github) along with a lightweight community portal called \textit{SciDash}.
+The overall goal of this work is to help scientists generate and examine tables like the one shown to Figure 1, where the relative validity of a set of models having a common scope can be determined by examining scores produced by a suite of validation tests constructed from experimental data.
+
 %Anecdotally, one of the greatest fears of new scientists is that a problem that they are working on has already been solved and published.
@@
 Unit tests serve this purpose for software, so we argue that the validation of neuroscience models by means of executable tests would be useful to scientists as a supplement the publication system.
-Test-driven methodologies have started to see success in neuroscience as well. 
-Modeling competitions in neuroscience, for example, are typically organized around a collection of simple validation criteria, implemented as executable tests. 
-These competitions continue to drive important advances and improve scientists' understanding of the relative merits of different models. 
-For example, the quantitative single neuron modeling competition (QSNMC) \cite{jolivet_quantitative_2008} investigates the complexity-accuracy tradeoff among reduced models of excitable membranes; 
-the ``Hopfield'' challenge \cite{hopfield_what_2000} tested techniques for generating neuronal network form given its function; 
-the Neural Prediction Challenge sought the best stimulus reconstructions, given neuronal activity (\url{http://neuralprediction.berkeley.edu}); the Diadem challenge is advancing the art of neurite reconstruction (\url{http://www.diademchallenge.org}); and
-examples from other subfields of biology abound (\url{http://www.the-dream-project.org}).
+% In addition to this open testing toolkit we also provide a library of tests for models of spiking single neurons.
+
+% Section 2.
+\section{Validation Testing with {SciUnit}}
+\begin{figure}[t]
+\small
+\begin{python}
+class PositionTest(sciunit.Test):
+    """Tests a planetary position model based on positions observed on day n given the positions in the n-1 previous days.
+    Metric: Standard p-value.
+    Parameters: 
+      obs_histories : list[list[position]]
+      obs_positions: list[position]"""
+	def __init__(self, obs_histories, obs_positions):
+		self.obs_histories = obs_histories
+		self.obs_positions = obs_positions
+
+	required_capabilities = [PredictsPlanetaryPosition]
+
+	def _judge(self, model):
+		predictions = []
+		for obs_history in self.obs_histories:
+			predictions.append(model.predict_next_pos(obs_history))
+		p = pooled_p_val(predictions, self.obs_positions)
+		return sciunit.PValue(p, related_data={
+			'obs_histories': obs_histories,
+			'obs_positions': obs_positions,
+			'predictions': predictions
+		})
+\end{python}
+\vspace{-5px}
+\caption{An example test class in \texttt{cosmounit}.}
+\label{fig:rate_test}
+\vspace{-15px}
+\end{figure}
+
+%Simple executable \emph{validation tests} that compute agreement between a model prediction and an experimental observation.  
+As a motivating example, we will begin by considering a community of early cosmologists recording and attempting to model observations of the planets visible in the sky, such as their position, velocity, orbital eccentricity and so on.
+One simple validation test might ask a model to predict planetary position on night $n+1$ given observations of its position on $n$ previous nights.
+Figure 2 shows how to implement a test, using  SciUnit, that captures this logic.
+
+Before explaining the details of this example, we point out that SciUnit is implemented in Python.
+Python is one of the most widely used languages in science today \cite{sanner1999python}.
+It supports calling into  many of the other popular languages, including R, MATLAB, C and Java, more cleanly than these languages support calling into Python (challenge 3).
+It is widely-recognized as being easy-to-read and its object system can be used to define abstract interfaces, which we leverage in support of challenge 2.
+
+A \textit{SciUnit} validation test is an {instance} of a Python class implementing the \verbx{sciunit.Test} abstract interface (line 1). 
+Here, the class \verbx{PositionTest} takes two \emph{parameters} in its constructor (constructors are named \verbx{\_\_init\_\_} in Python, lines 7-9). 
+The meaning of each parameter along with a description of the goodness-of-fit metric used by the test is documented on lines 2-6. 
+%For convenience, we also make use of functions provided by the popular NumPy\cite{numpy_url} and SciPy\cite{scipy_url} libraries, although these are not required by \textit{SciUnit}.  
+To create a \emph{particular} position test, we instantiate this class with particular planetary observations. 
+For example, the subset of cosmologists interested specifically in Saturn might instantiate a test by randomly chunking observations made about Saturn as follows:
+\begin{python}
+  h, p = randomly_chunk(saturn_obs_positions)
+  saturn_position_test = SpikeCountTest(h, p)
+\end{python}
+
+The class \verb|PositionTest| defines logic that is not specific to any particular planet, so it is contained in a package shared by all cosmologists called \verbx{cosmounit}, while the particular test above would be used in a test suite focused specifically on Saturn called \verbx{saturnsuite}.
+Both the common logic and specific suites are collaboratively developed by these overlapping research communities in source code repositories on Github (or another similar service).
+
+Classes that implement the \verbx{sciunit.Test} interface must contain a \verbx{_judge} method that receives a candidate \emph{model} as input and produces a \textit{score} as output. 
+To specify the interface between the test and the model, the test author provides a list of \emph{capabilities} in the \verbx{required_capabilities} attribute, seen on line 11 of Fig.
+\ref{fig:rate_test}.
+Capabilities are simply collections of methods that a test will need to invoke in order to receive relevant data, and are analogous to \emph{interfaces} in e.g. Java. 
+In Python, capabilities must be written as classes with unimplemented members. 
+The capability required by the test in Figure \ref{fig:rate_test} is shown in Figure \ref{fig:capability}. 
+In \textit{SciUnit}, classes defining capabilities are tagged as such by inheriting from \verbx{sciunit.Capability}.
+The test in Figure \ref{fig:rate_test} repeatedly uses this capability on line 16 to produce a position prediction for each observation.
+We assume that positions are represented in a standardized manner specified within \verbx{cosmounit}.
+A model is simply an object that implements capabilities (via Python's simple inheritance mechanism).
+
+The remainder of the \verbx{_judge} method compares the model predictions to observed data to produce a pooled $p$ value.
+The method returns an instance of \verbx{sciunit.PValue}, a subclass of \texttt{sciunit.Score} that has been included with \textit{SciUnit} due to its generality.
+In addition to the $p$-value itself, the returned score object also contains metadata, via the \verbx{related_data} parameter, for  scientists who may wish to examine the result in more detail later.
+This illustrates some key differences between unit testing, which would simply produce a boolean result, and our conception of scientific validation testing (challenge 1).
+A score must induce an ordering, so that the table shown in Figure 1 can be sorted along its columns, and it can optionally specify a normalization scheme so the cells can be color-coded (not shown).
+A test can be manually executed using the \verbx{judge} method:
+\begin{python}
+score = saturn_position_test.judge(kepler_sat)
+\end{python}
+
+This method proceeds by first checking that the provided model implements all required capabilities before calling the test's \verbx{_judge} method to produce a score.
+A reference to the test and model are added to the score for convenience (accessible via the \verbx{test} and \verbx{model} attributes, respectively).
+
+Alongside this position test, we could also provide a number of other test classes in \verbx{cosmounit} and instantiate them with data about aspects of Saturn's motion (e.g.
+it's velocity)  to produce a comprehensive suite in \verbx{saturnsuite}.
+\begin{python}
+saturn_motion = sciunit.TestSuite([saturn_position_test, saturn_velocity_test, ...])
+\end{python}
+Like a single test, a test suite is capable of judging one or more models. The result is a score matrix much like the one diagramed in Fig. \ref{fig:sciunit_overview}.
+\begin{python}
+sm_matrix = saturn_motion.judge([copernicus_sat, ptolemy_sat, newton_sat, kepler_sat])
+\end{python}
+A test suite requires the union of the capabilities required by the tests it contains.
+A model that performs well across tests in such a suite could reasonably claim (e.g.
+to reviewers) that it is a coherent, valid model of Saturn's motion.
+As new data is collected, new tests can be added to the suite.
+However, because the interface between the test and the model remains the same (only the data parameterizing the tests changes), the table can be updated automatically.
+Similarly, when a new model is developed, it can immediately be evaluated against all known data that has been encoded as a test by simply exposing its predictions via the capabilities the test requires.
+%They then defined a combined metric favoring models that broadly succeeded at meeting these criteria, to produce an overall ranking.
+%Such combined criteria are simply validation tests that invoke other tests to produce a result.
+
+%In many cases, models require no modifications to take the new tests because the same type of model output is being requested.
+%We discuss visualization of results in Secs. \ref{sec:scidash_activities} and \ref{sec:scidash_visualization}.
+
+\begin{figure}
+\begin{python}
+class PredictsPlanetaryPosition(sciunit.Capability):
+  def predict_next_pos(self, history): 
+    """Takes a list of previous positions and produces the next position."""
+    raise NotImplementedError("Model does not implement capability.")
+\end{python}
+\caption{An example capability specifying a single required method (used by the test in Figure \ref{fig:rate_test}).}
+\label{fig:capability}
+\vspace{-10px}
+\end{figure}
+
+%\begin{figure}
+%\begin{python}
+%class TrainSpikeCountFromCurrent(sciunit.Capability):
+%  def train_with_currents(self, currents, counts):
+%    """Takes a list of numpy arrays containing current stimulus (in nA) and
+%    observed spike counts. Model parameters should be adjusted based on this
+%    training data."""
+%    raise NotImplementedError("Model does not implement capability.")
+%\end{python}
+%\caption{Another capability specifying a training protocol (not used by the test in Figure \ref{fig:rate_test}).}
+%\label{fig:training}
+%\vspace{-15px}
+%\end{figure}
+
+%\subsection{Models}
+%Capabilities are \emph{implemented} by models.
+In \textit{SciUnit}, models are instances of Python classes that inherit from \verbx{sciunit.Model}.
+Like tests, the class itself represents a family of models, parameterized by the arguments of the constructor.
+A particular model is an instance of such a class.
+%
+%Figure \ref{fig:simple_model} shows how to write a simple {family} of models, \verbx{LinearModel}, that implement the capability in Fig.
+\ref{fig:capability} as well as another capability shown in Fig.
+\ref{fig:training}, which we will discuss below.
+%Models in this family generate a spike count by applying a linear transformation to the mean of the provided input current.
+The family is parameterized by the scale factor and the offset of the transformation, both scalars.
+%To create a \emph{particular} linear model, a modeler can provide particular parameter values, just as with test families:
+%\begin{python}
+%CA1_linear_model_heuristic = LinearModel(3.0, 1.0)
+%\end{python}
+%Here, the parameters to the model were picked by the modeler heuristically, or based on externally-available knowledge. 
+%An alternative test design would add a training phase where these parameters were fit to data using the capability shown in Fig. \ref{fig:training}. 
+%This test could thus only be used for those models for which parameters can be adjusted without human involvement. 
+%Whether to build a training phase into the test protocol is a choice left to each test development community. 
+%
+%Fig. \ref{fig:rate_test} does not include a training phase. 
+%If training data is externally available, models that nevertheless do implement a training capability (like \verb|LinearModel|) can simply be trained explicitly by calling the capability method just like any other Python method:
+%\begin{python}
+%CA1_linear_model_fit = LinearModel()
+%CA1_linear_model_fit.train_with_currents(CA1_training_in, CA1_training_out)
+%\end{python}
+%\begin{figure}
+%\begin{python}
+%class LinearModel(sciunit.Model, SpikeCountFromCurrent, 
+%    TrainSpikeCountFromCurrent):
+%  def __init__(self, scale=None, offset=None): 
+%    self.scale, self.offset = scale, offset
+%    
+%  def spike_count_from_current(self, input):
+%    return int(self.scale*numpy.mean(input) + self.offset)
+%
+%  def train_with_currents(self, currents, counts):
+%    means = [numpy.mean(c) for c in currents]
+%    [self.offset, self.scale] = numpy.polyfit(means, counts, deg=1)    
+%\end{python}
+%\caption{A model that returns a spike count by applying a linear transformation to the mean input current.
+The parameters can be provided manually or learned from data provided by a test or user (see text).}
+%\label{fig:simple_model}
+%\vspace{-15px}
+%\end{figure}
+%
+\section{Collaborative Workflow}
+The design just described has been purposefully left as simple as possible, in pursuit of challenge 4.
+Despite its simple design, it captures the essential elements of the scientific model validation process and satisfies the four challenges we laid out.
+SciUnit can be used by individual scientists to organize their workflows, but because science, and in particular, model validation is a collaborative process, we have also described the intended use of SciUnit within a collaborative workflow, mediated by a social coding tool like Github.
+
+More directly: we anticipate common testing logic (e.g.
+\verbx{PositionTest}), modeling logic and capabilities being collaboratively developed by larger communities in less specialized repositories like \verbx{cosmounit}.
+Individuals and small groups will then parameterize these tests and models with data they have gathered, as well as data known from the literature and data contained in existing data repositories to create test suites in repositories like \verbx{saturnsuite}.
+The interface between data collection software and existing data representation standards and tests will be mediated by bridge logic also contained in repositories like \verbx{cosmounit}.
+% This workflow is meant to evoke the \emph{test-driven development} methodology that is often used in software projects.
+
+Statisticians wishing to promote new validation metrics can simply for a $X$\verbx{unit} repository and implement new test logic.
+By reusing the same interfaces as previous tests used, these new metrics can immediately be used to validate or invalidate a large body of existing models, providing valuable data for use in convincing the community to adopt these into the mainstream repositories.
+Similarly, investigators who wish to de-emphasize or emphasize different quantities can fork $X$\verbx{suite} repository and add or remove tests.
+
+To help organize these repositories, a simple collaborative portal sitting above Github called \textit{SciDash} is currently under development (\url{http://scidash.org/}).
+SciDash consists of an organized, collaboratively filtered listing of $X$\verbx{unit} and $X$\verbx{suite} repositories.
+That is, it is a tool to help scientists find the most popular repositories in their research area, to facilitate the development of community standards.
+SciDash also supports extracting summaries of tests, models, scores and related data from suite repositories to generate hyperlinked score tables and documentation.
+This facilitates exploratory workflows.
+To modify a suite (e.g.
+by adding the model a scientist is developing to it), a scientist can simply fork the repository.
+SciDash automatically discovers public forks of repositories that it is already indexing.
+
+\section{Discussion}
+Academic peer review and code review are similar, but software developers typically supplement human review with test-driven methodologies.
+Such methodologies may benefit scientific communities as well, but several unique  constraints make reusing existing testing frameworks and processes difficult.
+We describe a core testing framework that addresses these issues  by using an existing, widely-adopted language and designing a flexible, simple framework that captures the domain-specific structure of scientific model validation.
+We then describe, by example, how the various components can be organized into software repositories and collaboratively maintained using social coding tools to support existing scientific practices.
+We end by outlining a collaboratively filtered portal that sits atop these tools and facilitates exploratory analyses and repository discovery.
+
+While we discuss a toy example based on planetary movement here, we have applied this framework to more realistic problems in neurobiology and have bridged SciUnit to large-scale neuroinformatics projects (we anticipate publishing a description of these case studies shortly).
+Much future work remains to be done to investigate whether these tools are truly usable and useful to scientific communities, to further develop the community workflow and infrastructure and to develop a range of realistic case studies.
+Nevertheless, we believe that identifying the synergies between testing practices and model validation, and describing the basic tooling, represents a novel contribution to the literature on software testing.
+The core SciUnit framework has been fully developed and is available at http://sciunit.scidash.org/.
+SciDash is under active development, and is currently capable of basic forms of the operations described in Sec.
+3.
+%In particular, we consider modeling competitions in science as a form of test-driven validation.
+For example, the quantitative single neuron modeling competition (QSNMC) \cite{jolivet_quantitative_2008} investigates the complexity-accuracy tradeoff among reduced models of excitable membranes, and
+%examples from other subfields of biology abound (e.g.
+\url{http://www.the-dream-project.org}).
+These competitions typically use \emph{ad hoc} infrastructure.
+A common framework could be used to make designing and continuously operating such contests more feasible.
+In related work that
+
+%Modeling competitions in neuroscience, for example, are typically organized around a collection of simple validation criteria, implemented as executable tests. 
+%These competitions continue to drive important advances and improve scientists' understanding of the relative merits of different models. 
 % \begin{figure}%{r}{0.6\textwidth}
@@
 
-Each of these examples has leveraged \emph{ad hoc} infrastructure to support test generation. 
-While the specific criteria used to evaluate models varies widely between disciplines in neuroscience, the underlying test-driven methodology has many common features that could be implemented once. 
-Recognizing this, we developed a discipline-agnostic framework for developing scientific validation test suites called \textit{SciUnit} (\url{http://www.sciunit.org}). 
-Here we describe \textit{NeuronUnit\-}, which builds upon \textit{SciUnit}, allowing neuroscientists to build \textit{SciUnit} tests that validate neurophysiology models against electrophysiological data.
-We provide a concrete example pipeline, showing how models described using \text{NeuroML} and provided freely by the \textit{Open Source Brain Project} (OSB, \cite{gleeson_open_2012}, \url{http://www.opensourcebrain.org}) can be tested in fully automated fashion using published, curated data available through the \textit{NeuroElectro Project} (Neuroelectro, \cite{tripathy_neuroelectro:_2012}, \url{http://neuroelectro.org}), leveraging facilities from the \textit{NeuroTools} library (\url{http://neuralensemble.org/NeuroTools}) to extract relevant features of model output.
-This is summarized in Figure \ref{fig:sciunit_overview}, which shows the relationships between the layers described here.
-% In addition to this open testing toolkit we also provide a library of tests for models of spiking single neurons.
+%Each of these examples has leveraged \emph{ad hoc} infrastructure to support test generation. 
+%While the specific criteria used to evaluate models varies widely between disciplines in neuroscience, the underlying test-driven methodology has many common features that could be implemented once. 
+%Recognizing this, we developed a discipline-agnostic framework for developing scientific validation test suites called \textit{SciUnit} (\url{http://www.sciunit.org}).
 
-% Section 2.
-\section{Validation Testing with \textit{SciUnit}}
-\subsection{Example: The Quantitative Single Neuron Modeling Competition}
-\begin{figure}[t]
-\small
-\begin{python}
-class SpikeCountTest(sciunit.Test):
-  """Tests spike counts produced in response to several current stimuli against observed means and standard deviations.
-
-  goodness of fit metric: Computes p-values based on a chi-squared test statistic, and pools them using Fisher's method.
-  parameters:
-    inputs: list of numpy arrays containing input currents (nA)
-    means, stds: lists of observed means and standard deviations, one per input
-  """
-  def __init__(self, inputs, means, stds):
-    self.inputs, self.means, self.stds = inputs, means, stds
-
-  required_capabilities = [SpikeCountFromCurrent]
-
-  def _judge(self, model):
-    inputs, means, stds = self.inputs, self.means, self.stds
-    n = len(inputs)
-    counts = numpy.empty((n,))
-    for i in xrange(n):
-      counts[i] = model.spike_count_from_current(inputs[i])
-    chisquared = sum((counts-means)**2 / means) # An array of chi-squared values.  
-    p = scipy.stats.chi2.cdf(chisquared,n-1) # An array of p-values.  
-    pooled_p = sciunit.utils.fisherp(p_array) # A pooled p-value.  
-    return sciunit.PValue(pooled_p, related_data={
-      "inputs": inputs, "counts": counts, "obs_means": means, "obs_stds": stds
-    })
-\end{python}
-\vspace{-5px}
-\caption{An example single neuron spike count test class implemented using \textit{SciUnit}.
-Because this implementation contains logic common to many different systems, \textit{NeuronUnit} was developed to provide a simpler means to deliver it (see Sec.
-\ref{sec:neuronunit}).}
-\label{fig:rate_test}
-\vspace{-15px}
-\end{figure}
-
-%Simple executable \emph{validation tests} that compute agreement between a model prediction and an experimental observation.  
-We first illustrate the form of a generic example \textit{SciUnit} test suite that could be used in neurophysiology. 
-Suppose we have collected data from an experiment where current stimuli (measured in pA) are delivered to neurons in some brain region, while the somatic membrane potential of each stimulated cell (in mV) is recorded and stored.
-A model claiming to capture this cell type's membrane potential dynamics must be able to accurately predict a variety of features observed in these data.
-
-One simple validation test would ask candidate models to predict the number of action potentials (a.k.a.
-spikes) generated in response to a stimulus (e.g.
-white noise), and compare these \emph{spike count} predictions to the distribution observed in repeated experimental trials using the same stimulus.
-For data of this type, goodness-of-fit can be measured by first calculating a p-value from a chi-squared statistic for each prediction and then combining these p-values using Fisher's method \cite{fisher_statistical_1925}.
-
-Alongside this \emph{spike count test}, we might also specify a number of other tests capturing different features of the data  to produce a more comprehensive suite. 
-For data of this sort, the QSNMC defined 17 other validation criteria in addition to one based on the overall spike count, capturing features like spike latencies (SL), mean subthreshold voltage (SV), interspike intervals (ISI) and interspike minima (ISM) that can be extracted from the data \cite{jolivet_quantitative_2008}.
-They then defined a combined metric favoring models that broadly succeeded at meeting these criteria, to produce an overall ranking. 
-Such combined criteria are simply validation tests that invoke other tests to produce a result.
-
-%In many cases, models require no modifications to take the new tests because the same type of model output is being requested.
-\subsection{Implementing a Validation Test in \textit{SciUnit}}
-Fig. \ref{fig:rate_test} shows how a scientist can implement spike count tests such as the one described above using \textit{SciUnit}. 
-A \textit{SciUnit} validation test is an {instance} (i.e. an object) of a Python class implementing the \verbx{sciunit.Test} interface (cf. line 1). 
-Here, we show a class \verbx{SpikeCountTest} taking three \emph{parameters} in its constructor (constructors are named \verbx{__init__} in Python, lines 9-10). 
-The meaning of each parameter along with a description of the goodness-of-fit metric used by the test is documented on lines 4-7. 
-%For convenience, we also make use of functions provided by the popular NumPy\cite{numpy_url} and SciPy\cite{scipy_url} libraries, although these are not required by \textit{SciUnit}.  
-To create a \emph{particular} spike count test, we instantiate this class with particular experimental observations. 
-For example, given observations from hippocampal CA1 cells (not shown), we can instantiate a test as follows:
-\begin{python}
-  CA1_sc_test = SpikeCountTest(CA1_inputs, CA1_means, CA1_stds)
-\end{python}
-We emphasize the crucial distinction between the \textit{class} \verbx{SpikeCountTest}, which defines a \emph{parameterized family} of validation tests, and the particular \textit{instance} \verbx{CA1_sc_test}, which is an individual validation test because the necessary parameters, derived from data, have been provided.
-As we will describe below, we expect communities to build repositories of such families capturing the criteria used in their subfields of neuroscience so that test generation for a particular system of interest will often require simply instantiating a previously-developed family with particular experimental parameters and data.
-For single-neuron test families like \verbx{SpikeCountTest}, we have developed such a library, called \textit{NeuronUnit} (\url{http://github.com/scidash/neuronunit}) (Sec. \ref{sec:neuronunit}).
-
-Classes that implement the \verbx{sciunit.Test} interface must contain a \verbx{_judge} method that receives a candidate \emph{model} as input and produces a \textit{score} as output. 
-To specify the interface between the test and the model (that is, to specify an appropriate scope), the test author provides a list of \emph{capabilities} in the \verbx{required_capabilities} attribute, seen on line 12 of Fig.
-\ref{fig:rate_test}.
-Capabilities are simply collections of methods that a test will need to invoke in order to receive relevant data, and are analogous to \emph{interfaces} in e.g.
-Java (\url{http://docs.oracle.com/javase/tutorial/java/concepts/interface.html}).
-In Python, capabilities are written as classes with unimplemented members. 
-The capability required by the test in Fig. \ref{fig:rate_test} is shown in Fig. \ref{fig:capability}. 
-In \textit{SciUnit}, classes defining capabilities are tagged as such by inheriting from \verbx{sciunit.Capability}.
-The test in Figure \ref{fig:rate_test} uses this capability on line 19 to produce a spike count prediction for each input current.
-
-The remainder of the \verbx{_judge} method implements the goodness-of-fit metric described above, returning an instance of \verbx{sciunit.PValue}, a subclass of \verbx{sciunit.Score} that is included with \textit{SciUnit}.
-In addition to the $p$-value itself, the returned score object also contains metadata, via the \verbx{related_data} parameter, for  scientists who may wish to examine the result in more detail later.
-In this case we save the input currents, the model outputs and the observed means and standard deviations (line 24). 
-%We discuss visualization of results in Secs. \ref{sec:scidash_activities} and \ref{sec:scidash_visualization}.
-
-\begin{figure}
-\begin{python}
-class SpikeCountFromCurrent(sciunit.Capability):
-  def spike_count_from_current(self, input): 
-    """Takes a numpy array containing current stimulus (in nA) and
-    produces an integer spike count. Can be called multiple times."""
-    raise NotImplementedError("Model does not implement capability.")
-\end{python}
-\caption{An example capability specifying a single required method (used by the test in Figure \ref{fig:rate_test}).}
-\label{fig:capability}
-\vspace{-10px}
-\end{figure}
-
-\begin{figure}
-\begin{python}
-class TrainSpikeCountFromCurrent(sciunit.Capability):
-  def train_with_currents(self, currents, counts):
-    """Takes a list of numpy arrays containing current stimulus (in nA) and
-    observed spike counts. Model parameters should be adjusted based on this
-    training data."""
-    raise NotImplementedError("Model does not implement capability.")
-\end{python}
-\caption{Another capability specifying a training protocol (not used by the test in Figure \ref{fig:rate_test}).}
-\label{fig:training}
-\vspace{-15px}
-\end{figure}
-
-\subsection{Models}
-Capabilities are \emph{implemented} by models.
-In \textit{SciUnit}, models are instances of Python classes that inherit from \verbx{sciunit.Model}.
-Like tests, the class itself represents a family of models, parameterized by the arguments of the constructor.
-A particular model is an instance of such a class.
-
-Figure \ref{fig:simple_model} shows how to write a simple {family} of models, \verbx{LinearModel}, that implement the capability in Fig.
-\ref{fig:capability} as well as another capability shown in Fig.
-\ref{fig:training}, which we will discuss below.
-Models in this family generate a spike count by applying a linear transformation to the mean of the provided input current.
-The family is parameterized by the scale factor and the offset of the transformation, both scalars.
-To create a \emph{particular} linear model, a modeler can provide particular parameter values, just as with test families:
-\begin{python}
-CA1_linear_model_heuristic = LinearModel(3.0, 1.0)
-\end{python}
-Here, the parameters to the model were picked by the modeler heuristically, or based on externally-available knowledge. 
-An alternative test design would add a training phase where these parameters were fit to data using the capability shown in Fig. \ref{fig:training}. 
-This test could thus only be used for those models for which parameters can be adjusted without human involvement. 
-Whether to build a training phase into the test protocol is a choice left to each test development community.
-
-Fig. \ref{fig:rate_test} does not include a training phase. 
-If training data is externally available, models that nevertheless do implement a training capability (like \verb|LinearModel|) can simply be trained explicitly by calling the capability method just like any other Python method:
-\begin{python}
-CA1_linear_model_fit = LinearModel()
-CA1_linear_model_fit.train_with_currents(CA1_training_in, CA1_training_out)
-\end{python}
-\begin{figure}
-\begin{python}
-class LinearModel(sciunit.Model, SpikeCountFromCurrent, 
-    TrainSpikeCountFromCurrent):
-  def __init__(self, scale=None, offset=None): 
-    self.scale, self.offset = scale, offset
-
-  def spike_count_from_current(self, input):
-    return int(self.scale*numpy.mean(input) + self.offset)
-
-  def train_with_currents(self, currents, counts):
-    means = [numpy.mean(c) for c in currents]
-    [self.offset, self.scale] = numpy.polyfit(means, counts, deg=1)    
-\end{python}
-\caption{A model that returns a spike count by applying a linear transformation to the mean input current.
-The parameters can be provided manually or learned from data provided by a test or user (see text).}
-\label{fig:simple_model}
-\vspace{-15px}
-\end{figure}
-
-\subsection{Executing Tests} A test is executed against a model using the \verbx{judge} method:
-\begin{python}
-score = CA1_sc_test.judge(CA1_linear_model_heuristic)
-\end{python}
-
-This method proceeds by first checking that the provided model implements all required capabilities.
-It then calls the test's \verbx{_judge} method to produce a score.
-A reference to the test and model are added to the score for convenience (accessible via the \verbx{test} and \verbx{model} attributes, respectively), before it is returned.
-
-\subsection{Test Suites and Score Matrices} A collection of tests intended to be run on the same model can be put together to form a test suite.
-The following is a test suite that could be used for a simplified version of the QSNMC, as described above:  
-\begin{python}
-CA1_suite = sciunit.TestSuite([CA1_sc_test, CA1_sl_test, CA1_sv_test, CA1_isi_test, CA1_ism_test])
-\end{python}
-Like a single test, a test suite is capable of judging one or more models. The result is a score matrix much like the one diagramed in Fig. \ref{fig:sciunit_overview}.
-\begin{python}
-CA1_matrix = CA1_suite.judge([CA1_linear_model_heuristic, CA1_linear_model_fit])
-\end{python}
-A simple summary of the scores in a score matrix can be printed to the console or  visualized by other tools, such as the web application \textit{SciDash} described in Sec. \ref{sec:scidash}.
+\section*{Acknowledgements}
+We thank Sharon Crook, Shreejoy Tripathy and Padraig Gleeson for their support of this project.  
+The work was supported in part by grant R01MH081905 from the National Institute of Mental Health. 
+The content is solely the responsibility of the authors and does not necessarily represent the official views of the National Institutes of Health.
 
@@
 \bibliographystyle{IEEEtran}
-\bibliography{../research}
+\bibliography{../research,references,references_crook,urls,references_ja}
+%\bibliography{references,references_crook,urls,references_ja}
+
+%\bibliographystyle{unsrt}
+%\bibliography{references,references_crook,urls,references_ja}
+%\listoftodos
+%\end{document}
+
 %

### 2013-12-22 att-icfp14/att-icfp14.tex
Pasted ESOP paper into ICFP14

--- 2013-10-12 esop14/att-esop14.tex
@@
-\documentclass{llncs}
+\documentclass[9pt,preprint]{sigplanconf}
 \usepackage{amsmath}
-\usepackage{llncsdoc}
 \usepackage{amssymb} 
-%\usepackage{amsthm}
+\usepackage{amsthm}
 \usepackage{ stmaryrd }
@@
 \begin{document}
-\mainmatter  % start of an individual contribution
-
-% first the title is needed
+
+\conferenceinfo{-}{-} 
+\copyrightyear{-} 
+\copyrightdata{[to be supplied]}
+
+%\titlebanner{{\tt \textcolor{Red}{{\small Under Review -- distribute within CMU only.}}}}        % These are ignored unless
+%\preprintfooter{Distribute within CMU only.}   % 'preprint' option specified.
+
 \title{Active Typechecking and Translation: A Safe Language-Internal Extension Mechanism}
 
-% a short form should be given in case it is too long for the running head
-\titlerunning{Lecture Notes in Computer Science: Authors' Instructions}
-
-% the name(s) of the author(s) follow(s) next
-%
-% NB: Chinese authors should write their first names(s) in front of
-% their surnames. This ensures that the names appear correctly in
-% the running heads and the author index.
-%
-\author{Cyrus Omar%
-\and Jonathan Aldrich}
-%
-\authorrunning{Lecture Notes in Computer Science: Authors' Instructions}
-% (feature abused for this document to repeat the title also on left hand pages)
-
-% the affiliations are given next; don't give your e-mail address
-% unless you accept that it will be published
-\institute{Carnegie Mellon University, Pittsburgh, PA 15213, USA\\
-\texttt{\{comar,aldrich\}@cs.cmu.edu}
-}
-
-%
-% NB: a more complex sample for affiliations and the mapping to the
-% corresponding authors can be found in the file "llncs.dem"
-% (search for the string "\mainmatter" where a contribution starts).
-% "llncs.dem" accompanies the document class "llncs.cls".
-%
-
-\toctitle{Lecture Notes in Computer Science}
-\tocauthor{Authors' Instructions}
+\authorinfo{~}{~}{~}
+%\authorinfo{Cyrus Omar\and Jonathan Aldrich}
+ %          {School of Computer Science\\
+  %          Carnegie Mellon University}
+   %        {\{comar, jonathan.aldrich\}@cs.cmu.edu}
+
 \maketitle
-
 \begin{abstract}
@@
 AT\&T leverages type-level computation, typed compilation techniques and a form of type abstraction to enable library-based implementations of a variety of primitives over a flexible  grammar in a safe manner.
-%In a language supporting AT\&T, primitive types families are represented uniformly at the type level.
-Each type is equipped with operators, defined using type-level functions, that are invoked specifically during the type-checking and translation of operations relevant to that type.
-%We discuss three points in the design space: (1) a simple judgmental construction of AT\&T designed to distill the essential concepts and admit formal safety theorems, (2) a "real" language called Ace that we use to demonstrate the expressive power of AT\&T in practice, and (3) a language design called Birdie that lifts and extends CoqÍs rich system of dependent types into the type-level, allowing extension developers to fully prove the correctness of extensions statically.
-\keywords{extensible languages; active libraries; typed compilation; type-level computation; type abstraction}
 \end{abstract}
 
+%\category{D.3.2}{Programming Languages}{Language Classifications}[Extensible Languages]
+%\category{D.3.4}{Programming Languages}{Processors}[Compilers]
+%\category{F.3.1}{Logics \& Meanings of Programs}{Specifying and Verifying and Reasoning about Programs}[Specification Techniques]
+%\keywords
+%type-level computation, typed compilation
 \section{Motivation}
@@
 This approach is often effective because modern {general-purpose} abstraction mechanisms, like inductive datatypes and object systems, are highly expressive.  
-For example, the Delite framework leverages Scala's powerful general-purpose mechanisms to enable a number of useful \emph{embedded domain-specific languages}.
-%\cite{delite}
+For example, the Delite framework leverages Scala's powerful general-purpose mechanisms to enable a number of useful \emph{embedded domain-specific languages} \cite{delite}.
 Unfortunately, there remain some situations of interest where general-purpose abstractions fall short. 
-For instance, it is difficult to adequately encode advanced type systems in terms of the simpler rules governing general-purpose abstractions (e.g.
+For instance, it is difficult to adequately encode advanced type systems in terms of the simpler rules that govern general-purpose abstractions (e.g.
 reasoning about units of measure requires built-in language support in F\# \cite{conf/cefp/Kennedy09}).
 Even if a full encoding is possible, it may not be useful if it is overly verbose or unnatural, or if the error messages are overly abstract.
-For example, regular expressions encoded using inductive datatypes are often considered overly verbose, so most functional languages support them via strings, which is less safe.
-Finally, general-purpose abstractions are implemented in a uniform manner.
-Domain knowledge is not easily applied to eliminate overhead or perform optimizations, and implementations designed for typical application workloads may not be satisfactory in parts of a program where performance is a key criteria, particularly when targeting heterogeneous hardware platforms (e.g.
+For example, regular expressions encoded using inductive datatypes are quite verbose, so most functional languages support them via strings, which is less safe.
+Finally, general-purpose abstractions are implemented in a uniform manner, meaning domain-specific heuristics cannot be applied to eliminate overhead or perform optimizations, and implementations designed for typical application workloads may not be satisfactory in parts of a program where performance is a key criteria, such as when targeting heterogeneous hardware platforms (e.g.
 programmable GPUs) and distributed computing resources.
-Using variants of familiar abstractions backed by specialized implementation strategies can be quite useful in these cases.
 
@@
 
-Researchers or domain experts who run into situations like these, where more direct control over a language's semantics and implementation are needed, have little choice today but to realize new abstractions by creating a new language of some form.
+Researchers or domain experts who run into situations like these, where more direct control over a language's semantics and implementation are needed, have little choice today but to realize new abstractions by creating a new language in some way.
 They might develop a new standalone language from scratch, modify an implementation of an existing language, or use tools like compiler generators, DSL frameworks and language workbenches \cite{fowler2010domain}.
@@
 The increasing sophistication and ease-of-use of these tools have led to calls for a {\it language-oriented approach} to software development, where different components of an application are written in different specialized languages \cite{journals/stp/Ward94}.
-Indeed, a number of software ecosystems are now explicitly designed to support many different languages, both general-purpose and domain-specific, atop a common intermediate language.
+Indeed, a number of software ecosystems are now designed explicitly to support many different languages, both general-purpose and domain-specific, atop a common intermediate language.
 The Java virtual machine (JVM), the Common Language Infrastructure (CLI) and LLVM are prominent examples of such ecosystems.
@@
 This compromises both verifiability and performance.
-For%requires the development of an interoperability layer for every pair of DSLs.
+%Forrequires the development of an interoperability layer for every pair of DSLs.
 Moreover, this approach exposes the internals of an implementation to clients, making the abstraction awkward to work with and causing code breakage when implementation details change.
@@
 %As an example, F\#'s type system prevents \lstinline{null} values from occurring within data structures, but because it's type system is not available when calling into F\# code from another language, like C\#, run-time null checks must still be included in the implementation.
-\begin{figure}[t]
+\begin{figure*}[t]
 \vspace{-15px}
 \includegraphics[scale=0.415]{approaches.pdf}
-\vspace{-30px}
 \caption{(a) With the language-oriented approach, different primitive abstractions are packaged into separate languages that extend and target a common intermediate language (e.g.
@@
 Clients can import and use these abstractions directly whenever needed.}
-\vspace{-10px}
 \label{approaches}
-\end{figure}
+\end{figure*}
 
@@
 %Developers of libraries need only determine whether they are appropriate for their domain, without also considering whether these constructs can be exposed in terms of abstractions appropriate to client code.
-As a result, clients can granularly import any necessary primitive constructs when using code that relies on them, and thus achieve full safety, ease-of-use and performance due to the absence of wrappers and glue code.
+As a result, clients can granularly import any necessary primitive constructs when using code that relies on them, and thereby achieve full safety, ease-of-use and performance.
 Providers of components thus need only consider whether primitives that they use are appropriate for their domain, without also considering whether their code might be used in a context where these primitives are not otherwise appropriate.
@@
 
-In this paper, we introduce a language-internal extensibility mechanism called \emph{active typechecking and translation} (AT\&T) that allows developers to introduce and implement the logic governing new primitive type families and operators from within libraries.
+In this paper, we introduce a language-internal extensibility mechanism called \emph{active typechecking and translation} (AT\&T) that allows developers to introduce and implement the logic governing new primitive type and operator families from within libraries.
 We argue that this can be accomplished by enriching the type-level language, rather than introducing a separate metalanguage into the system. 
 To make this proposal concrete, we begin by introducing a simple core calculus, called \atlam~(for the ``actively-typed lambda calculus''), in Section \ref{atlam}. 
-This calculus uses type-level computation of higher kind, along with techniques borrowed from the typed compilation literature and a form of type abstraction that ensures that the implementation details of an extension are not externally visible to guarantee the safety of the language, the decidability of typechecking and compilation and composability of extensions, as we outline in Section \ref{safety}.
+This calculus uses type-level computation of higher kind, along with techniques borrowed from the typed compilation literature and a form of type abstraction that ensures that the implementation details of an extension are not externally visible to guarantee the safety of the language, the decidability of typechecking and compilation and non-interference of extensions, as we outline in Section \ref{safety}.
 In Section \ref{examples}, we suggest that despite these constraints, this mechanism is expressive enough to admit, within libraries, a number of general-purpose and domain-specific abstractions that normally require built-in language support.
@@
 
-A more appropriate and useful place for extensions like this is directly within libraries, alongside abstractions that do not require this level of control.
-To enable this, the language must allow for the introduction new primitive type families, like \lstinline{Prod}, operators, like \lstinline{Pair}, \lstinline{PrL} and \lstinline{PrR}, and associated typechecking and translation logic, without allowing so much control that the reliability of the system suffers.
+A more appropriate and useful place for extensions like this is directly within libraries, alongside abstractions that can be adequately implemented in terms of existing primitive abstractions.
+To enable this, the language must allow for the introduction new primitive type families, like \lstinline{Prod}, operators, like \lstinline{Pair}, \lstinline{PrL} and \lstinline{PrR}, and associated typechecking and translation logic.
 When encountering these new operators in expressions, the compiler must effectively  hand control over typechecking and translation to the appropriate user-defined logic.
@@
 In fact, we will show that \textbf{the most natural place for type system extensions is within the type-level language}.
-The intuition is that extensions to a language's static semantics will need to manipulate types as values at compile-time.
+The intuition is that extensions to a statically-typed language's semantics will need to manipulate types as values at compile-time.
 Many languages already allow users to write type-level functions for various reasons, effectively supporting this notion of types as values at compile-time (see Sec.
@@
 The type-level language is often constrained by its own type system (where the types of type-level values are called \emph{kinds} for clarity) that prevents type-level functions from causing problems during compilation.
-This is precisely the structure that a distinct extension layer would have, and so it is quite natural to unify the two.
+This is precisely the structure that a distinct extension layer would have, and so it is quite natural to unify the two, as we will show in this work.
 
@@
 Variables $x$ are used in expressions and internal terms and are distinct from type-level variables, $\tvar{t}$.
-Names $\fvar{Fam}$ are family names (we assume that unique family names can be generated by some external mechanism) and $\opvar{op}$ are operator names.
+Names $\fvar{Fam}$ are type family names (we assume that globally unique type family names can be generated by some external mechanism) and $\opvar{op}$ are operator family names.
 $\tstr{str}$ denotes string literals, $\iintlit$ denotes integer literals and $\oplus$ stands for binary operations over integers.
@@
 A program, $\rho$, consists of a series of declarations followed by an expression.
-Declarations can be either bindings of type-level terms to type-level variables using \textsf{def} or a declaration of a new primitive type family using \textsf{family}.
-Expressions, $e$, can be either variables, lambdas, or applications of operators, and are ultimately given meaning by translation to a typed internal language, with terms $\gamma$ and types $\sigma$.
-This language has been chosen, for simplicity, to be a variant of Plotkin's PCF with primitive integers and products, but in practice would include other constructs consistent with its role as a high-level intermediate language.
+Declarations can be either bindings of type-level terms to type-level variables using \textsf{def} or a primitive type family declared using \textsf{family}, which can contain implementations of one or more operator families, $\theta$.
+Expressions, $e$, can be either variables, lambdas, or applications of operators.
+%, and are ultimately given meaning by translation to a typed internal language, with terms $\gamma$ and types $\sigma$. 
+%This language has been chosen, for simplicity, to be a variant of Plotkin's PCF with primitive integers and products, but in practice would include other constructs consistent with its role as a high-level intermediate language.
 
@@
 The type-level language also includes other kinds of terms: type-level functions, lists (required by our mechanism), integers, strings and products for the sake of our examples (see Sec.
-\ref{termination} for a discussion on other acceptable kinds of type-level data) and constructs for developing extensions -- denotations and reified internal terms and types -- which we will discuss in the sections below.
+\ref{safety} for a discussion on other acceptable kinds of type-level data) and constructs for developing extensions -- denotations and reified internal terms and types -- which we will discuss in the sections below.
+
+All expressions are given meaning by translation to a typed internal language.
+This language has been chosen, for simplicity, to be a variant of Plotkin's PCF with primitive integers and products, but in practice would include other constructs consistent with its role as a high-level intermediate language.
 The grammar of internal terms, $\gamma$, and internal types, $\sigma$, also includes special forms containing type-level terms; these are used for developing extensions and during compilation and will be erased before compilation ends.
@@
 \caption{G\"odel's T in \atlam, used to calculate 2+2.
-The statics we implement here are shown in the Appendix.
-Simple helper functions for working with lists ($\tvar{empty}$, $\tvar{pop}$, $\tvar{pop\_final}$) and types ($\tvar{check\_type}$),  described below, are given in the appendix.}
+Helper functions for working with lists ($\tvar{empty}$, $\tvar{pop}$, $\tvar{pop\_final}$) and types ($\tvar{check\_type}$),  described below, are given in the appendix.
+We will refer to this program as $\rho_{\text{nat}}$ in the text.}
 \label{example}
@@
 We define the typechecking and translation logic for the operators associated with this family (\opvar{z}, \opvar{s} and \opvar{rec}) on lines \ref{z}-\ref{lastop} and use these to define a $plus$ function on line \ref{add} and compute 2+2.
+We will refer back to this figure as we describe each construct below.
 
@@
 \bibliography{../research}
-\appendix
-\section{Appendix}
-\verb|http://www.cs.cmu.edu/~comar/att-esop14-appendix.pdf|.
+
+%\acks
+%The author is grateful to Jonathan Aldrich, Robert Bocchino and anonymous referees for their useful suggestions.
+This work was funded by the DOE Computational Science Graduate Fellowship under grant number DE-FG02-97ER25308.
+%Acknowledgments, if needed.
+
 \end{document}

### 2014-01-09 thesis-proposal/background.tex
continuing work on proposal

@@
 \subsubsection{Background: Active Libraries}\label{alibs}
-The term \emph{active libraries} was first used by Veldhuizen et al.
+The term \emph{active libraries} was first introduced by Veldhuizen et al.
 \cite{atthesis, atpaper} to describe ``libraries that take an active role in compilation, rather than being passive collections of subroutines''.
@@
 A prominent example in the literature is Blitz++, a library that uses C++ template metaprogramming to optimize compound operations on vectors and matrices by eliminating intermediate allocations \cite{blitzpp}.
-Although this and several other interesting optimizations have been shown possible by this technique, its expressiveness is fundamentally limited because template expansion allow for only the substitution of compile-time constants into pre-written code, and template metaprograms are notoriously difficult to read, write, debug and reason about.
+Although this and several other interesting optimizations have been shown possible by this technique, its expressiveness is fundamentally limited because template expansion allows for only the substitution of compile-time constants into pre-written code, and template metaprograms are notoriously difficult to read, write, debug and reason about \cite{Robison:2001:IEC:376656.376751}.
 %They are, however, composable, because templates are associated with a single function or class, and  the only metatheoretic issue is that template expansion can be non-terminating.
 
-More powerful compile-time \emph{term rewriting mechanisms} integrated into some languages can also be used for optimization, as well as for inserting specialized error checking and reporting\- logic and extending the language with powerful domain-specific abstractions.
+More powerful compile-time \emph{term rewriting mechanisms} integrated into some languages can also be used for optimization, as well as for introducing specialized error checking and reporting\- logic and extending the language with new abstractions.
 These mechanisms are highly expressive because they allow users to programmatically manipulate syntax trees directly, but they suffer from problems of composability and safety.
-Macros, such as those in MetaML \cite{metaml}, Template Haskell \cite{template_haskell} and others\todo{reference other macro systems}, can take full control over all the code that they enclose, so there is no way to guarantee that nested macros will compose as intended.
-Outer macros can neglect to invoke or  manipulate the output of inner macros in ways that can cause conflicts or weaken important guarantees.
-Once data escapes a macro's scope, there is no way to rely on the guarantees and features that were available within its scope -- the output of a macro is a value in the underlying language, so the underlying language is all that governs it (a problem related to the interoperability problem of Section \ref{motivation}).
-It can also be difficult to reason about the semantics of code when a number of enclosing macros may be manipulating it.
+Compile-time macros, such as those in MetaML \cite{metaml}, Template Haskell \cite{template_haskell} and Scala \cite{scala-macros}, can take full control over all the code that they enclose, so outer macros can neglect to invoke or  manipulate the output of inner macros, sometimes inadvertantly.
+Once a value escapes a macro's scope, there is no way to rely on the guarantees and features that were available within its scope, because the output of a macro is simply a value in the underlying language (a problem fundamentally related to the compatibility problem described in Section \ref{motivation}).
+Thus, macros can be used to automate code generation, but not to truly extend the semantics of a language.
+It can also be difficult to reason about the semantics of code when any number of enclosing macros may be manipulating it, and to build tools that can handle them.
 
 Some rewriting systems go beyond the block scoping of macros.
-Xroma (pronounced ``Chroma''), for example, is designed around active libraries and allows users to insert custom rewriting and error checking passes into the compiler from within libraries \cite{xroma}.
+Xroma (pronounced ``Chroma''), for example, is designed around active libraries and allows users to insert custom rewriting passes into the compiler from within libraries \cite{xroma}.
 Similarly, the derivative of Haskell implemented by the Glasgow Haskell Compiler (GHC) allows libraries to define custom compile-time term rewriting logic if an appropriate flag is passed in \cite{haskell-rewrite}.
-In both cases, the user-defined logic can dispatch on arbitrary patterns of code throughout the component or program the extension is active in, so these mechanisms are highly expressive and avoid some of the difficulties of block-scoped macros.
+In both cases, the user-defined logic can dispatch on arbitrary patterns of code throughout the component or program the extension is activated within, so these mechanisms are highly expressive and avoid some of the difficulties of block-scoped macros.
 But libraries containing such global rewriting logic cannot be safely composed because two different libraries may attempt to rewrite the same piece of code differently.
@@
 
-Another example of an active library approach to extensibility is SugarJ \cite{sugarj}.
-SugarJ allows libraries to extend the base syntax of Java in a nearly-arbitrary manner, and these extensions are imported transitively throughout a program.
-SugarJ libraries are thus also not safely composable.
+Another example of an active library approach to extensibility is SugarJ \cite{sugarj} and other languages in the Sugar* family \cite{sugarstar}, like SugarHaskell.
+These languages permit libraries to extend the base syntax of the core language in a nearly-arbitrary manner, and these extensions are imported transitively throughout a program.
+These extensions are also not safely composable.
 For example, a library that defines a literal syntax for HTML would conflict with another that defines a literal syntax for XML because they define differing semantics for some of the same syntactic forms.

### 2014-01-09 thesis-proposal/problem.tex
continuing work on proposal

@@
 \section{Motivation}\label{motivation}
-Designing and implementing a programming language together with its supporting tools (collectively, a \emph{programming system}) that has a sound theoretical foundation, helps users identify and fix errors as early as possible, supports a natural programming style, and that performs well across diverse problem domains and hardware platforms remains a grand challenge in computing.
-Due to the increasing diversity and complexity of modern problem domains and hardware platforms, it has become clear that no small  collection of general-purpose primitives backed by a simple toolchain will be able to fully satisfy these criteria in all cases.
-Instead,  researchers will continue to provide specialized abstractions, notations,  type systems, implementation strategies, optimizations, run-time systems and tool support (collectively, specialized \emph{features}) in order to meet these growing challenges.
+Specifying and implementing a programming language together with its supporting tools (collectively, a \emph{programming system}) that has sound theoretical foundations, helps users identify and fix errors as early as possible, supports a natural programming style, and performs well across diverse problem domains and hardware platforms remains a grand challenge in computing.
+Due to the increasing diversity and complexity of modern problem domains and hardware platforms, it has become clear that no small  collection of general-purpose constructs backed by a simple, all-purpose implementation can fully satisfy these criteria in all cases.
+Instead,  researchers and domain experts (collectively, \emph{providers}) continue to develop  specialized notations,  type systems, semantics, implementation strategies, optimizations, run-time systems and tools (collectively, \emph{features})  designed to address these modern challenges in different circumstances.
 %, researchers and domain experts must continue to design and develop, in a principled manner, specialized notations, type systems, optimizations, run-time systems and tools (collectively, \emph{features}) to support developers as they work in increasingly complex problem domains and on modern hardware systems.
@@
 
-Ideally, providers would develop and distribute new features as libraries, so that developers could granularly choose those that best satisfy the constraints of their problem domains.
-Unfortunately, this is often infeasible, because libraries are nearly exclusively vehicles for specifying the run-time behaviors of a program, but providing these kinds of features  requires or benefits from some measure of control over the compile-time or edit-time semantics of the programming system.
-That is, from the perspective of a library, the language's semantics are fixed in advance, the compiler and run-time system are ``black box'' implementations of these fixed semantics, and the other tools, like code editors and debuggers, operate according to domain-agnostic protocols also based exclusively on these fixed semantics.
-As a result, providers of new system features must take \emph{language-external approaches}.
-We will argue that such approaches have several fundamental problems, and that taking them has led to an unnecessary gap between research and practice.
-In place of these approaches, we will advocate for \emph{language-internal extensibility mechanisms}, and show how, by organizing new features around types and constraining them appropriately, such mechanisms can be made both safe and highly expressive.
+Ideally, a provider would develop and distribute such new features orthogonally, as libraries, so that client developers could granularly choose those that best satisfy their needs.
+Unfortunately this is often infeasible because libraries are vehicles for specifying a program in terms of a monolithic collection of existing features.
+From the perspective of a library, the language's syntax, type system and semantics are fixed in advance, the compiler and run-time system are ``black box'' implementations of this fixed language, and the other tools, like code editors and debuggers, operate according to domain-agnostic protocols also based only on a fixed language specification.
+As a result, providers of new system features must typically take \emph{language-external approaches}, sometimes designing a new language altogether.
+We will argue that such approaches are problematic, and that taking them has led to an unnecessary gap between research and practice.
+In place of these approaches, we will advocate for \emph{language-integrated extensibility mechanisms}, and show how, by organizing new features around {types} and imposing certain constraints, these mechanisms can be made both safe and highly expressive.
+We call a type that introduces new features an \emph{active type} and programming systems organized around active types \emph{actively-typed programming systems}.
 %But taking a \emph{language-internal approach} to implementing a feature is the most practical.
@@
 To make the issue concrete, let us begin with a simple example.
-\emph{Regular expressions} are a widely-used mechanism for finding patterns in unstructured and semi-structured strings (when formal grammars are not appropriate) \cite{regex}.
-If a programming system included full compile-time and edit-time support for regular expressions, it might provide features like:
+\emph{Regular expressions} are a widely-used mechanism for finding patterns in semi-structured strings (e.g.
+DNA sequences) \cite{Thompson:1968:PTR:363347.363387}.
+A programming system that included full compile-time and edit-time support for regular expressions might simultaneously provide features like these:
 
 \begin{enumerate}
-\item \textbf{built-in syntax for pattern literals} so that malformed patterns result in intelligible \emph{compile-time} parsing errors, motivated by the frequency of run-time errors relating to pattern syntax in Java \cite{regex-type-system} and other languages that do not have this feature.
-\item \textbf{type checking logic} that ensures that key constraints related to regular expressions are not violated, such as that out-of-bounds group indices are not used \cite{regex-type-system} and that only values with correct types are spliced into regular expressions.
-When a type error is found, the error message should again be intelligible.
-\item \textbf{translation logic} that partially or fully compiles regular expressions into the efficient internal representation that will be used by the regular expression matching engine at run-time.
-In most languages, this compilation step occurs at run-time, even if the pattern is fully known at compile-time, thereby introducing latency into programs.
-If the developer is not careful, regular expressions used repeatedly in a program might be needlessly re-compiled on each use.
+\item \textbf{Built-in syntax for pattern literals} (e.g.
+\cite{perlre}) so that malformed patterns result in intelligible \emph{compile-time} parsing errors.
+In languages lacking such literals, run-time errors relating to pattern syntax can occur, even in well-tested, deployed code \cite{spishak2012type}.
+\item \textbf{Typechecking logic} that ensures that key constraints related to regular expressions are not violated at compile-time:
+	\begin{enumerate}
+	\item only appropriate values are spliced into regular expressions, to avoid splicing errors and injection attacks \cite{owasp2013}
+	\item out-of-bounds backreferences are not used \cite{spishak2012type}
+	\item strings known to be in the language of a regular expression are given a type that tracks this information and ensures that string manipulation operations do not inadvertently lead to a string that is not in the assumed language \cite{fulton-thesis}.
+	\end{enumerate}
+When a type error is found, an intelligible error message is provided.
+\item \textbf{Elaboration logic} that partially or fully compiles known regular expressions into the efficient internal representation that will be used by the regular expression matching engine (e.g., a finite automata \cite{Thompson:1968:PTR:363347.363387}) ahead of time.
+In most languages, this compilation step occurs at run-time, even if the pattern is fully known at compile-time, thereby introducing performance overhead into programs.
+If the developer is not careful to cache compiled representations, regular expressions used repeatedly in a program might be needlessly re-compiled on each use.
 %By performing this step ahead-of-time, these dangers can be avoided.
-\item \textbf{editor-integrated tooling} for interactively testing patterns against example strings, quickly referring to documentation, searching databases of common patterns and other domain-specific edit-time facilities.
+\item \textbf{Editor services} for quickly testing regular expression patterns against test strings,  referring to documentation or searching databases of common patterns (e.g. \cite{IntelliJRegexp}).
 \end{enumerate}
@@
 Instead, libraries generally provide support for regular expressions by leveraging  general-purpose constructs.
-Unfortunately, it is impossible to fully encode the syntax and the specialized static and dynamic semantics described above in terms of standard general-purpose notations and abstractions.
-Library providers have thus needed to compromise, typically by representing regular expressions using strings and deferring parsing, typechecking and translation to run-time, which introduces performance overhead and can lead to unanticipated run-time errors (as shown in \cite{regex-type-system}) and security vulnerabilities (due to injection attacks when splicing is used, for example).
-Similarly, edit-time features, as described above, are rarely integrated into editors, and never in a way that facilitates their discovery and use directly when the developer is manipulating regular expressions.
-In most cases, tools must be discovered independently and accessed externally (for example, via a browser), making both their development and use less common and more awkward than necessary \cite{active-code-completion} (see Section \ref{acc}).
+Unfortunately, it is impossible to fully encode the syntax and the specialized static and dynamic semantics described above in terms of general-purpose notations and abstractions.
+Library providers have thus needed to compromise.
+The typical choice is to have clients provide regular expressions as strings, deferring parsing, typechecking and elaboration to run-time, which introduces performance overhead and can lead to unanticipated run-time errors (as shown in \cite{spishak2012type}) and security vulnerabilities (due to injection attacks when user inputs are spliced into patterns, for example).
+Similarly, tools are rarely integrated into editors, and even more rarely in a way that facilitates their discovery and use directly when the developer is manipulating regular expressions.
+Tools that must be discovered independently and accessed externally (for example, via a browser) are used less frequently  and can be more awkward than necessary \cite{Campbell:2008:DRT:1636642.1636651, Omar:2012:ACC:2337223.2337324}, leading to lower productivity.
 
@@
 \vspace{-20px}
-\caption{\small (a) With the language-external approach, novel constructs are packaged into separate languages.
-Users can only safely and naturally call into languages if the interface uses common constructs and an interoperability layer has been developed (the \emph{interoperability problem}).
+\caption{\small (a) With the language-external approach, novel features are packaged into separate languages and tools.
+Clients of an API can only safely and naturally use it if the interface uses common constructs (the \emph{compatibility problem}).
 (b) With the active library approach, there is one extensible host language and the compile-time and edit-time logic governing novel constructs is expressed within libraries.
@@
 
-When the compile-time or edit-time semantics of a system must be modified to fully realize a new feature, as in the example above, providers often take a \emph{language-external approach}, either by developing a new or derivative programming system (often centered around a so-called \emph{domain-specific language} \cite{dsl}) or by extending an existing system by some mechanism that is not part of the language itself, such as an extension mechanism supported by a {particular} compiler\footnote{Note that compilers that modify or allow modification of  the semantics of their base language, rather than simply implementing meaning-preserving optimizations, should be considered as pernicious means for creating new languages.
-Many compilers, including \texttt{gcc}, \texttt{GHC} and \texttt{SML/NJ}, are guilty of this sin, meaning that some programs that seem to be written in C, Haskell or Standard ML are actually written in tool-specific derivatives of those languages.
-Language-internal mechanisms do not lead to such fragmentation.} or other tool.
+When the primitive behaviors of a system must be extended to fully realize a new feature, as in the example above, providers typically take a \emph{language-external approach}, either by developing a new or derivative programming system (often centered around a so-called \emph{domain-specific language (DSL)} \cite{fowler2010domain}) or by extending an existing system by some mechanism that is not part of the language itself, such as an extension mechanism for a {particular} compiler\footnote{Compilers that modify, or allow modification of, the semantics of their base language, rather than simply permitting meaning-preserving optimizations, should be considered a pernicious means for creating new languages.
+Many ``practical'' compilers, including \texttt{gcc}, \texttt{GHC} and \texttt{SML/NJ}, are guilty of this sin, meaning that some programs that seem to be written in C, Haskell or Standard ML are actually written in tool-specific derivatives of those languages.
+Language-internal mechanisms do not lead to such fragmentation.} or other tool\todo{relate to regex example and update figure}.
 %This latter method couples the semantics of the feature to the implementation details of a particular tool.
@@
 
-Unfortunately, when providers of new features take language-external approaches, it causes problems for clients.
-Features cannot be adopted individually, but instead are only available coupled to a collection of other often-unrelated  features (the incidental features included in a newly-designed language or tool, or the existing features of the particular tool that was externally extended).
-This makes adoption more costly when these other features are either not appropriate or insufficiently developed, or when the features bundled with a different language or tool are simultaneously desirable.
-For example, although evidence suggests that developers prefer language-integrated parallel programming abstractions to library-based implementations if all else is equal \cite{langvslib}, library-based implementations are more widely adopted because parallel programming languages privilege only a few abstractions, even though parallel programming is rarely the only concern relevant to a component or application.
+Unfortunately, when providers of new features take language-external approaches, it causes problems for clients that hinder adoption.
+Features cannot be adopted individually, but instead are only available coupled to a collection of other unrelated  features (e.g.
+the incidental features of a newly-designed language or tool, or of the particular tool that was extended).
+This makes adoption more costly when these incidental features are  not appropriate or insufficiently developed, or when the features bundled with a different language or tool are simultaneously desirable.
+For example, although evidence suggests that developers prefer language-integrated parallel programming abstractions to library-based implementations if all else is equal \cite{langvslib}, library-based implementations are more widely adopted because ``parallel programming languages'' provide full support only for a few chosen parallel programming abstractions along with a small collection of less-developed general-purpose abstractions.
+This is problematic because different parallel abstractions are more appropriate in different situations \cite{ecoopstudy13} and parallel programming support is rarely the only concern relevant to client developers outside of a classroom setting.
 Regular expression support, for example, may be simultaneously desirable when processing large amounts of textual data in parallel.
@@
 
-Even when various concerns can be separated into different components, each written in a suitable language (often called the \emph{language-oriented approach} to software development \cite{language-oriented}), the interfaces between these components remains an issue.
-The specialized primitives particular to one language cannot always be safely and naturally expressed in terms of those available in another, so building a program out of components written in a variety of different languages is difficult or impossible whenever these primitives are exposed at interface boundaries.
-We refer to this fundamental issue as the \emph{interoperability problem}.
+Various concerns can sometimes be separated into different components, each written using a suitable language and tooling.
+This has been called the \emph{language-oriented approach} to software development \cite{language-oriented} and is supported by \emph{language workbenches} \cite{workbenches} and domain-specific language (DSL) frameworks \cite{fowler10}.
+Unfortunately, this is insufficient: the interfaces between components written in different languages remains an issue.
+The specialized constructs particular to one language (e.g.
+futures in a parallel programming language) cannot always be safely and naturally expressed in terms of those available in another (e.g.
+a general-purpose language), so building a program out of components written in a variety of different languages is difficult whenever these are exposed at interface boundaries.
+Tool support is also lost when calling into a different language.
+We call this fundamental issue the \emph{compatibility problem}.
 
-One strategy often taken by programming language designers to partially address the interoperability problem is to  target an established language or bytecode, such as the Java Virtual Machine (JVM) bytecode or LLVM IR, and support a superset of its constructs.
-Scala \cite{scala} and F\# \cite{fsharp} are two prominent examples of general-purpose languages that have taken this approach, and many domain-specific language frameworks also rely on this strategy (e.g.
-Delite \cite{delite}).
+One strategy often taken by proponents of the language-oriented approach to partially address the compatibility problem is to  target an established intermediate language, such as the Java Virtual Machine (JVM) bytecode\todo{mention functional IL} and support a superset of its constructs.
+Scala \cite{scala} and F\# \cite{fsharp} are examples of prominent general-purpose languages that have taken this approach, and most DSL frameworks also rely on this strategy.
 This only enables full interoperability in one direction (Figure \ref{approaches}a).
 While calling into the common language becomes straightforward, calls in the other direction, or between the languages sharing the common target, are still restricted by the semantics of the common language.
+
 If new languages only include constructs that can already be expressed safely and reasonably naturally in the common language, then this approach can work well. 
@@
 The F\# type system also includes support for checking that units of measure are used correctly \cite{fsharpunits}, but this specialized invariant is left completely unchecked at language boundaries.
-In Scala, interfaces built around traits that have default method implementations are difficult to implement from Java or other JVM languages and the workaround can break if the trait is modified \cite{scalatraitinterop}.
+%In Scala, traits that have default method implementations are difficult to implement from Java or other JVM languages and the workaround can break if the trait is modified \cite{scalatraitinterop}. 
 In some cases, desirable features must be omitted entirely due to concerns about  interoperability.
-The module system in F\#, for example, is substantially simpler than that in its predecessor, OCaml, despite F\# otherwise aiming to maintain compatibility, due to the need for bidirectional interoperability between F\# and other CLI languages.
+F\#, for example, aims to maintain compatibility with ML, but due to the need for bidirectional interoperability with CLI languages, it cannot support modules or functors.
 %\end{itemize}
 
-\subsection{Active Libraries}
-For these reasons, we argue that taking a language-external approach to realizing a new feature, by packaging it into a new or derivative language or tool, should be considered harmful and avoided whenever possible.
-The goal of the research being proposed here is to fundamentally reorganize the core components of the programming system so that such language-external approaches are less frequently necessary, by designing \emph{language-internal extension mechanisms} that give developers control over edit-time and compile-time behaviors that have previously been defined externally and thus, in the case of popular languages, become available only subject to the approval of slow-moving governing committees\footnote{One might compare today's monolithic programming systems to  {centrally-planned} economies, whereas extensible\- systems more closely resemble modern market economies.}\todo{read Arch D.
-Robison.
-Impact of economics on compiler optimization.}.
-Specifically, we will show how control over \textbf{parsing}, \textbf{typechecking}, \textbf{translation} (the first stage of compilation) and \textbf{code completion} can be delegated to user-defined logic distributed in {libraries}, as illustrated in Figure \ref{approaches}b.
-Such libraries are called \emph{active libraries} because, rather than being passive consumers of features already available in the system, they contain logic invoked by the system during development or compilation to provide new domain-specific features  \cite{active-libraries}.
-Features implemented within active libraries are imported as needed by the clients of libraries that rely on them, unlike features implemented by language-external means.
+\subsection{Language-Integrated Approaches}
+We argue that, due to these compatibility problems, taking a language-external approach to realizing a new feature should be considered harmful and avoided whenever possible.
+The goal of the research being proposed here is to design \emph{language-integrated extension mechanisms} that give developers control over edit-time and compile-time behaviors that have previously been defined centrally\footnote{One might compare today's monolithic programming systems to  {centrally-planned} economies, whereas extensible\- systems more closely resemble modern market economies.} so that such language-external approaches are less frequently necessary.
+More specifically, we will show how control over \textbf{parsing}, \textbf{typechecking}, \textbf{elaboration} and \textbf{code completion} can be delegated to user-defined logic distributed in {libraries}, as illustrated in Figure \ref{approaches}b.
+Such libraries are called \emph{active libraries} because, rather than being passive clients of features already available in the system, they contain logic invoked by the system during development or compilation to provide new features  \cite{active-libraries}.
+Features implemented within active libraries are imported as needed by the clients of libraries that rely on them, unlike features implemented by language-external means, avoiding the compatibility problem.
 
-Some critical issues having to do with {safety} must be overcome before library-based extension mechanisms can be introduced into a programming system, because if too much control over such core aspects of the system is given  to developers, the system may become unreliable.
+We must proceed with caution, however: critical issues having to do with {safety} must be overcome before library-based extension mechanisms can be introduced into a programming system, because if too much control over such core aspects of the system is given  to developers, the system may become unreliable.
 %For example, an extension could weaken important metatheoretic guarantees previously provided by the system. 
@@
 For example, if two active libraries defined differing semantics for the same syntactic form, the issue would only manifest itself when both libraries were imported somewhere within the same program.
-These kinds of safety issues have plagued previous attempts at language-internal extensibility.% To prevent them, our mechanisms will organize  extension logic around types to guarantee that extensions are both safe in isolation and also safely composable in any combination.
+These kinds of safety issues have plagued previous attempts to design language-integrated extensibility mechanisms.% To prevent them, our mechanisms will organize  extension logic around types to guarantee that extensions are both safe in isolation and also safely composable in any combination.
 

### 2014-01-09 thesis-proposal/listing4.py

@@ -1,15 +1,10 @@
-from ace.OpenCL import OpenCL, get_global_id, 
-  global_ptr, double, int
-
-@OpenCL.fn
-def map(input, output, f):
-    gid = get_global_id(0)
-    output[gid] = f(input[gid])
+from listing3 import map
+from ace.OpenCL import gptr, double, int
 
 @OpenCL.fn
 def add5(x):
     return x + 5
 
-A = GlobalPtrType(double); B = GlobalPtrType(int)
-map_add5_dbl = map.compile(A, A, add5.ace_type)
-map_add5_int = map.compile(B, B, add5.ace_type)
+D = gptr(double); I = gptr(int); A = add5.ace_type
+map_add5_dbl = map.compile(D, D, A)
+map_add5_int = map.compile(I, I, A)

### 2014-01-09 thesis-proposal/listing7.py

@@ -1,0 +1,10 @@
+from ace.OpenCL import OpenCL
+import ace.astx as astx
+
+plus = OpenCL.fn.from_str("""
+def plus(a, b):
+    return a + b
+""")
+
+add5_ast = astx.specialize(plus.ast, b=5)
+add5 = OpenCL.fn.from_ast(add5_ast)

### 2014-01-09 thesis-proposal/listing8.py

@@ -1,0 +1,16 @@
+import numpy as np
+import ace.OpenCL.bindings as cl
+from listing3 import map
+from listing4 import add5
+
+cl.ctx = cl.Context.for_device(0, 0)
+
+input = np.ones((1024,))
+d_in = cl.to_device(input)
+d_out = cl.alloc(like=d_in)
+
+map(d_in, d_out, add5, 
+    global_size=d_in.shape, local_size=(128,))
+
+out = cl.from_device(d_out)
+assert (out == input + 5).all()

### 2014-01-09 thesis-proposal/listing9.py

@@ -1,8 +1,9 @@
-import ace, ace.astx as astx
+import ace
 
-class GlobalPtrType(ace.Type):
-  def __init__(self, target_type):
+class PtrType(ace.Type):
+  def __init__(self, target_type, addr_space):
     self.target_type = target_type
+    self.addr_space = addr_space
         
   def resolve_Subscript(self, context, node):
     slice_type = context.resolve(node.slice)
@@ -14,8 +15,11 @@ class GlobalPtrType(ace.Type):
   def translate_Subscript(self, context, node):
     value = context.translate(node.value)
     slice = context.translate(node.slice)
-    return astx.copy_node(node, 
+    return ace.copy_node(node, 
       value=value, slice=slice,    
-      code=value.code + '[' + slice.code + ']')
-      
+      code=value.code + '[' + slice.code + ']')      
+  
   # ...
+  
+def gptr(target_type):
+    return PtrType(target_type, "__global")

### 2014-01-09 thesis-proposal/problem.tex
Updated figure

@@
 Similarly, tools are rarely integrated into editors, and even more rarely in a way that facilitates their discovery and use directly when the developer is manipulating regular expressions.
-Tools that must be discovered independently and accessed externally (for example, via a browser) are used less frequently  and can be more awkward than necessary \cite{Campbell:2008:DRT:1636642.1636651, Omar:2012:ACC:2337223.2337324}, leading to lower productivity.
+Tools that must be discovered independently and accessed externally (for example, via a browser) are used less frequently  and can be more awkward than necessary \cite{Campbell:2008:DRT:1636642.1636651, Omar:2012:ACC:2337223.2337324}, leading to lower productivity \cite{Omar:2012:ACC:2337223.2337324}.
 
@@
 \begin{center}
-\includegraphics[scale=0.45]{approaches.pdf}
+\includegraphics[scale=.45]{approaches.pdf}
 \end{center}

### 2014-01-09 thesis-proposal/approaches-crop.sh

@@ -1,0 +1,2 @@
+pdfcrop --margins "-3 -3 -3 -3" approaches.pdf approaches.pdf
+

### 2014-01-10 thesis-proposal/problem.tex
Proposal finalized through sec. 1.2

--- 2014-01-09 thesis-proposal/problem.tex
@@
 Due to the increasing diversity and complexity of modern problem domains and hardware platforms, it has become clear that no small  collection of general-purpose constructs backed by a simple, all-purpose implementation can fully satisfy these criteria in all cases.
-Instead,  researchers and domain experts (collectively, \emph{providers}) continue to develop  specialized notations,  type systems, semantics, implementation strategies, optimizations, run-time systems and tools (collectively, \emph{features})  designed to address these modern challenges in different circumstances.
+Instead,  researchers and domain experts (collectively, \emph{providers}) continue to develop  specialized notations,  static semantics, dynamic semantics, implementation strategies, optimizations, run-time systems and tools (collectively, \emph{features})  designed to address these modern challenges in different circumstances.
 %, researchers and domain experts must continue to design and develop, in a principled manner, specialized notations, type systems, optimizations, run-time systems and tools (collectively, \emph{features}) to support developers as they work in increasingly complex problem domains and on modern hardware systems.
@@
 From the perspective of a library, the language's syntax, type system and semantics are fixed in advance, the compiler and run-time system are ``black box'' implementations of this fixed language, and the other tools, like code editors and debuggers, operate according to domain-agnostic protocols also based only on a fixed language specification.
-As a result, providers of new system features must typically take \emph{language-external approaches}, sometimes designing a new language altogether.
+As a result, providers of new system features must often take \emph{language-external approaches}, sometimes designing a new language and its associated tools altogether.
 We will argue that such approaches are problematic, and that taking them has led to an unnecessary gap between research and practice.
-In place of these approaches, we will advocate for \emph{language-integrated extensibility mechanisms}, and show how, by organizing new features around {types} and imposing certain constraints, these mechanisms can be made both safe and highly expressive.
+In place of these approaches, we will advocate for \emph{language-integrated extensibility mechanisms}, and show how, by organizing new features around {types} enforcing certain constraints, these mechanisms can be made both safe and highly expressive.
 We call a type that introduces new features an \emph{active type} and programming systems organized around active types \emph{actively-typed programming systems}.
@@
 Instead, libraries generally provide support for regular expressions by leveraging  general-purpose constructs.
-Unfortunately, it is impossible to fully encode the syntax and the specialized static and dynamic semantics described above in terms of general-purpose notations and abstractions.
-Library providers have thus needed to compromise.
-The typical choice is to have clients provide regular expressions as strings, deferring parsing, typechecking and elaboration to run-time, which introduces performance overhead and can lead to unanticipated run-time errors (as shown in \cite{spishak2012type}) and security vulnerabilities (due to injection attacks when user inputs are spliced into patterns, for example).
-Similarly, tools are rarely integrated into editors, and even more rarely in a way that facilitates their discovery and use directly when the developer is manipulating regular expressions.
-Tools that must be discovered independently and accessed externally (for example, via a browser) are used less frequently  and can be more awkward than necessary \cite{Campbell:2008:DRT:1636642.1636651, Omar:2012:ACC:2337223.2337324}, leading to lower productivity.
+Unfortunately, it is impossible to fully encode the syntax and the specialized static and dynamic semantics described above in terms of general-purpose notations and abstractions like objects or inductive datatypes.
+Library providers have thus needed to compromise, typically by asking clients to provide regular expressions as strings, thereby deferring parsing, typechecking and elaboration to run-time, which introduces performance overhead and can lead to unanticipated run-time errors (as shown in \cite{spishak2012type}) and security vulnerabilities (due to injection attacks when user inputs are spliced into patterns, for example).
+Similarly, useful tools for working with regular expressions are rarely integrated into editors, and even more rarely in a way that facilitates their discovery and use directly when the developer is manipulating regular expressions.
+Tools that must be discovered independently and accessed externally are used less frequently  and can be more awkward than necessary \cite{Campbell:2008:DRT:1636642.1636651, Omar:2012:ACC:2337223.2337324}, leading to lower productivity \cite{Omar:2012:ACC:2337223.2337324}.
 
@@
 \begin{center}
-\includegraphics[scale=0.45]{approaches.pdf}
+\includegraphics[scale=.48]{approaches.pdf}
 \end{center}
 \vspace{-20px}
-\caption{\small (a) With the language-external approach, novel features are packaged into separate languages and tools.
-Clients of an API can only safely and naturally use it if the interface uses common constructs (the \emph{compatibility problem}).
-(b) With the active library approach, there is one extensible host language and the compile-time and edit-time logic governing novel constructs is expressed within libraries.
-If the extension mechanism guarantees the safety of arbitrary compositions of extensions, the necessary primitives can simply be imported by library clients as needed, and so interoperability is not a problem.}
+\caption{\small (a) When taking a language-external approach, new features are packaged together into separate languages and tools, causing problems with compatibility (described in the text).
+(b) When taking a language-integrated approach, there is one extensible host language and the compile-time and edit-time logic governing new constructs is expressed within ``active'' libraries.
+It is critical that the extension mechanism  guarantees that extensions cannot interfere with one another, so that the compatibility problem is avoided.}
 \label{approaches}
@@
 
-When the primitive behaviors of a system must be extended to fully realize a new feature, as in the example above, providers typically take a \emph{language-external approach}, either by developing a new or derivative programming system (often centered around a so-called \emph{domain-specific language (DSL)} \cite{fowler2010domain}) or by extending an existing system by some mechanism that is not part of the language itself, such as an extension mechanism for a {particular} compiler\footnote{Compilers that modify, or allow modification of, the semantics of their base language, rather than simply permitting meaning-preserving optimizations, should be considered a pernicious means for creating new languages.
+When the semantics or implementation of a system must be extended to fully realize a new feature, as in the example above, providers typically take a \emph{language-external approach}, either by developing a new or derivative programming system (supported by \emph{language workbenches} \cite{erdweg2013state},  \emph{DSL frameworks} \cite{fowler2010domain} or \emph{compiler generators}), or by extending an existing system by some mechanism that is not part of the language itself, such as an extension mechanism for a {particular} compiler\footnote{Compilers that modify, or allow modification of, the semantics of their base language, rather than simply permitting meaning-preserving optimizations, should be considered a pernicious means for creating new languages.
 Many ``practical'' compilers, including \texttt{gcc}, \texttt{GHC} and \texttt{SML/NJ}, are guilty of this sin, meaning that some programs that seem to be written in C, Haskell or Standard ML are actually written in tool-specific derivatives of those languages.
-Language-internal mechanisms do not lead to such fragmentation.} or other tool\todo{relate to regex example and update figure}.
-%This latter method couples the semantics of the feature to the implementation details of a particular tool.
+Language-internal mechanisms do not lead to such fragmentation.} or other tool.
+For example, a researcher interested in providing regexp-related features (let us refer to these collectively as \texttt{R}) might design a new language with built-in support for regular expressions, perhaps basing it on an existing language (containing some general-purpose constructs, \texttt{G}).
+A different researcher developing a new language-based parallel programming abstraction or implementation strategy (collectively, \texttt{P}) might  take the same strategy.
+A third researcher, developing an alternative parallel programming abstraction (collectively, \texttt{Q}) might again do the same.
+This results in a collection of distinct systems as diagrammed in Figure \ref{approaches}a.%This latter method couples the semantics of the feature to the implementation details of a particular tool.
 Because the use of one implementation entails a different semantics for the feature than another, the extended tool acts, \emph{de facto}, as a distinct system for our purposes.
 
-Unfortunately, when providers of new features take language-external approaches, it causes problems for clients that hinder adoption.
+Unfortunately, when providers of new features take language-external approaches like this, it causes  problems for clients.
 Features cannot be adopted individually, but instead are only available coupled to a collection of other unrelated  features (e.g.
-the incidental features of a newly-designed language or tool, or of the particular tool that was extended).
-This makes adoption more costly when these incidental features are  not appropriate or insufficiently developed, or when the features bundled with a different language or tool are simultaneously desirable.
-For example, although evidence suggests that developers prefer language-integrated parallel programming abstractions to library-based implementations if all else is equal \cite{langvslib}, library-based implementations are more widely adopted because ``parallel programming languages'' provide full support only for a few chosen parallel programming abstractions along with a small collection of less-developed general-purpose abstractions.
-This is problematic because different parallel abstractions are more appropriate in different situations \cite{ecoopstudy13} and parallel programming support is rarely the only concern relevant to client developers outside of a classroom setting.
-Regular expression support, for example, may be simultaneously desirable when processing large amounts of textual data in parallel.
+the incidental features of the newly-designed language and its associated tools).
+This makes adoption more costly when these incidental features are  not desirable or insufficiently developed, or when the features bundled with a different language or tool are simultaneously desirable.
+Evidence supports the prominence of this problem: developers prefer language-integrated parallel programming abstractions to library-based implementations if all else is equal \cite{cave2010comparing}, but library-based implementations are more widely adopted because ``parallel programming languages'' privilege only a few chosen  abstractions and implementation strategies.
+This is problematic because different parallel abstractions or implementation strategies are more appropriate in different situations \cite{Tasharofi:2013rc} and moreover,  parallel programming support is rarely the only concern relevant to client developers outside of a classroom setting.
+Regular expression support, for example, may be simultaneously desirable for processing large amounts of textual data in parallel, but using these features together in the same compilation unit when they have been implemented by language-external means is difficult or impossible.
+One must either use the system containing features \texttt{G+R} or \texttt{G+P}.
+There is no system containing both.
 %Similarly, a language and tools designed primarily to support regular expressions might make an interesting research project, but it would not be a suitable tool for writing large applications with more varied needs.
@@
 
-Various concerns can sometimes be separated into different components, each written using a suitable language and tooling.
-This has been called the \emph{language-oriented approach} to software development \cite{language-oriented} and is supported by \emph{language workbenches} \cite{workbenches} and domain-specific language (DSL) frameworks \cite{fowler10}.
-Unfortunately, this is insufficient: the interfaces between components written in different languages remains an issue.
+Different application concerns can sometimes be separated into different components, each written using a suitable language and tooling.
+This has been called the \emph{language-oriented approach} to software development \cite{journals/stp/Ward94}.
+Unfortunately, this is insufficient: the interface between components written in different languages remains an issue.
 The specialized constructs particular to one language (e.g.
@@
 Tool support is also lost when calling into a different language.
-We call this fundamental issue the \emph{compatibility problem}.
+We call this fundamental issue the \emph{client compatibility problem}: clients of a certain collection of features cannot always interface with clients of a different collection of features in a safe, performant and natural manner.
 
-One strategy often taken by proponents of the language-oriented approach to partially address the compatibility problem is to  target an established intermediate language, such as the Java Virtual Machine (JVM) bytecode\todo{mention functional IL} and support a superset of its constructs.
-Scala \cite{scala} and F\# \cite{fsharp} are examples of prominent general-purpose languages that have taken this approach, and most DSL frameworks also rely on this strategy.
-This only enables full interoperability in one direction (Figure \ref{approaches}a).
+One strategy often taken by proponents of the language-oriented approach to partially address the compatibility problem is to  target an established intermediate language, such as the Java Virtual Machine (JVM) bytecode, and support a superset of its constructs.
+Scala \cite{200464/IC} and F\# \cite{pickering2007foundations} are examples of prominent general-purpose languages that have taken this approach, and most DSL frameworks also rely on this strategy.
+As indicated in Figure \ref{approaches}a, this only enables true client compatibility in one direction.
 While calling into the common language becomes straightforward, calls in the other direction, or between the languages sharing the common target, are still restricted by the semantics of the common language.
@@
 For example, the type system of F\# guarantees that null values cannot occur within F\# data structures, but maintaining this important invariant still requires run-time checks because the typing rules of F\# do not apply when F\# code is called from other languages on the Common Language Infrastructure (CLI) like C\#.
-The F\# type system also includes support for checking that units of measure are used correctly \cite{fsharpunits}, but this specialized invariant is left completely unchecked at language boundaries.
+The F\# type system also includes support for checking that units of measure are used correctly \cite{syme2012expert, kennedy1994dimension}, but this specialized invariant is left completely unchecked at language boundaries.
 %In Scala, traits that have default method implementations are difficult to implement from Java or other JVM languages and the workaround can break if the trait is modified \cite{scalatraitinterop}. 
 In some cases, desirable features must be omitted entirely due to concerns about  interoperability.
-F\#, for example, aims to maintain compatibility with ML, but due to the need for bidirectional interoperability with CLI languages, it cannot support modules or functors.
+F\#, for example, is based on Ocaml, but due to the need for bidirectional interoperability with CLI languages, it cannot support advanced features like polymorphic variants, modules or functors \cite{ocaml-manual}.
 %\end{itemize}

### 2014-01-11 thesis-proposal/background.tex
Final Sec. 1

@@
 % !TEX root = omar-thesis-proposal.tex
-\subsubsection{Background: Active Libraries}\label{alibs}
+\subsubsection{Background}\label{alibs}
 The term \emph{active libraries} was first introduced by Veldhuizen et al.
-\cite{atthesis, atpaper} to describe ``libraries that take an active role in compilation, rather than being passive collections of subroutines''.
-The authors went on to suggest a number of concrete reasons libraries might benefit by being able to influence the programming system at compile-time or edit-time, including high-level program optimization, checking programs for correctness against specialized criteria, reporting domain-specific errors and warnings, and ``rendering domain-specific textual and non-textual program representations and for interacting with such representations'' (anticipating interactions between libraries and tools other than just the compiler.
-Our definition, given in Section \ref{motivation}, differs  from theirs, partly for this reason).
+\cite{activelibraries, active-libraries-thesis} to describe ``libraries that take an active role in compilation, rather than being passive collections of subroutines''.
+The authors suggested a number of reasons libraries might benefit by being able to influence the programming system at compile-time or edit-time, including high-level program optimization, checking programs for correctness against specialized criteria, reporting domain-specific errors and warnings, and ``rendering domain-specific textual and non-textual program representations and for interacting with such representations'' (anticipating interactions between libraries and tools other than just the compiler).
 
@@
 The first concrete realizations of active libraries, prompting the introduction of the term, were libraries that performed domain-specific program optimization at compile-time by exploiting language mechanisms that allow for limited compile-time computation.
-A prominent example in the literature is Blitz++, a library that uses C++ template metaprogramming to optimize compound operations on vectors and matrices by eliminating intermediate allocations \cite{blitzpp}.
-Although this and several other interesting optimizations have been shown possible by this technique, its expressiveness is fundamentally limited because template expansion allows for only the substitution of compile-time constants into pre-written code, and template metaprograms are notoriously difficult to read, write, debug and reason about \cite{Robison:2001:IEC:376656.376751}.
+A prominent example in the literature is Blitz++, a library that uses C++ template metaprogramming to optimize compound operations on vectors and matrices by eliminating intermediate allocations \cite{veldhuizen2000blitz++}.
+Although this and several other interesting optimizations have been shown possible by this technique, its expressiveness is fundamentally limited because template expansion allows for only the substitution of compile-time constants into pre-written code, and template metaprograms are notoriously difficult to read, write, debug and reason about (see discussion in \cite{Robison:2001:IEC:376656.376751}).
 %They are, however, composable, because templates are associated with a single function or class, and  the only metatheoretic issue is that template expansion can be non-terminating.
 
-More powerful compile-time \emph{term rewriting mechanisms} integrated into some languages can also be used for optimization, as well as for introducing specialized error checking and reporting\- logic and extending the language with new abstractions.
+More powerful and direct compile-time \emph{term rewriting mechanisms} available in some languages can also be used for optimization, as well as for introducing specialized error checking and reporting\- logic and extending a language with new abstractions.
 These mechanisms are highly expressive because they allow users to programmatically manipulate syntax trees directly, but they suffer from problems of composability and safety.
-Compile-time macros, such as those in MetaML \cite{metaml}, Template Haskell \cite{template_haskell} and Scala \cite{scala-macros}, can take full control over all the code that they enclose, so outer macros can neglect to invoke or  manipulate the output of inner macros, sometimes inadvertantly.
-Once a value escapes a macro's scope, there is no way to rely on the guarantees and features that were available within its scope, because the output of a macro is simply a value in the underlying language (a problem fundamentally related to the compatibility problem described in Section \ref{motivation}).
+For example, compile-time macros, such as those in MetaML \cite{Sheard:1999:UMS}, Template Haskell \cite{SheardPeytonJones:Haskell-02} and Scala \cite{scala-macros}, take full control over all the code that they enclose.
+This can be problematic, however, as outer macros can interfere with the functionality of inner macros.
+Moreover, once a value escapes a macro's scope, there is no way to rely on the guarantees and features that were available within its scope, because the output of a macro is simply a value in the underlying language (a problem fundamentally related to the compatibility problems described in Section \ref{motivation}).
 Thus, macros can be used to automate code generation, but not to truly extend the semantics of a language.
-It can also be difficult to reason about the semantics of code when any number of enclosing macros may be manipulating it, and to build tools that can handle them.
+It can also be difficult to reason about the semantics of code when any number of enclosing macros may be manipulating it, and to build tools that operate robustly in their presence.
 
-Some rewriting systems go beyond the block scoping of macros.
-Xroma (pronounced ``Chroma''), for example, is designed around active libraries and allows users to insert custom rewriting passes into the compiler from within libraries \cite{xroma}.
-Similarly, the derivative of Haskell implemented by the Glasgow Haskell Compiler (GHC) allows libraries to define custom compile-time term rewriting logic if an appropriate flag is passed in \cite{haskell-rewrite}.
+Some term rewriting systems replace the block scoping of macros with pattern-based dispatch.
+Xroma (pronounced ``Chroma''), for example, is designed around active libraries and allows users to insert custom rewriting passes into the compiler from within libraries \cite{activelibraries}.
+Similarly, the derivative of Haskell implemented by the Glasgow Haskell Compiler (GHC) allows libraries to define custom compile-time term rewriting logic if an appropriate flag is passed in \cite{jones2001playing}.
 In both cases, the user-defined logic can dispatch on arbitrary patterns of code throughout the component or program the extension is activated within, so these mechanisms are highly expressive and avoid some of the difficulties of block-scoped macros.
 But libraries containing such global rewriting logic cannot be safely composed because two different libraries may attempt to rewrite the same piece of code differently.
-It is also difficult to guarantee that such logic is correct and difficult for users to reason about code when simply importing a library can change the semantics of the program in a non-local manner.
+It is also difficult to guarantee that such logic is correct and difficult to reason about code when importing a library can change the semantics of the program in a non-local manner.
 % http://www.haskell.org/ghc/docs/latest/html/users_guide/rewrite-rules.html
 
-Another example of an active library approach to extensibility is SugarJ \cite{sugarj} and other languages in the Sugar* family \cite{sugarstar}, like SugarHaskell.
+Another example of an active library approach to extensibility is SugarJ \cite{erdweg2011sugarj} and other generated by Sugar* \cite{erdweg2013framework}, like SugarHaskell \cite{erdweg2012layout}.
 These languages permit libraries to extend the base syntax of the core language in a nearly-arbitrary manner, and these extensions are imported transitively throughout a program.

### 2014-01-11 thesis-proposal/problem.tex
Final Sec. 1

@@
 From the perspective of a library, the language's syntax, type system and semantics are fixed in advance, the compiler and run-time system are ``black box'' implementations of this fixed language, and the other tools, like code editors and debuggers, operate according to domain-agnostic protocols also based only on a fixed language specification.
-As a result, providers of new system features must often take \emph{language-external approaches}, sometimes designing a new language and its associated tools altogether.
+As a result, providers of new system features must often take \emph{language-external approaches}, sometimes designing a new system altogether.
 We will argue that such approaches are problematic, and that taking them has led to an unnecessary gap between research and practice.
@@
 In some cases, desirable features must be omitted entirely due to concerns about  interoperability.
-F\#, for example, is based on Ocaml, but due to the need for bidirectional interoperability with CLI languages, it cannot support advanced features like polymorphic variants, modules or functors \cite{ocaml-manual}.
+F\#, for example, is based on Ocaml, but due to the need for bidirectional interoperability with CLI languages, it cannot support features like polymorphic variants, modules or functors \cite{ocaml-manual}.
 %\end{itemize}
@@
 \subsection{Language-Integrated Approaches}
-We argue that, due to these compatibility problems, taking a language-external approach to realizing a new feature should be considered harmful and avoided whenever possible.
-The goal of the research being proposed here is to design \emph{language-integrated extension mechanisms} that give developers control over edit-time and compile-time behaviors that have previously been defined centrally\footnote{One might compare today's monolithic programming systems to  {centrally-planned} economies, whereas extensible\- systems more closely resemble modern market economies.} so that such language-external approaches are less frequently necessary.
-More specifically, we will show how control over \textbf{parsing}, \textbf{typechecking}, \textbf{elaboration} and \textbf{code completion} can be delegated to user-defined logic distributed in {libraries}, as illustrated in Figure \ref{approaches}b.
-Such libraries are called \emph{active libraries} because, rather than being passive clients of features already available in the system, they contain logic invoked by the system during development or compilation to provide new features  \cite{active-libraries}.
-Features implemented within active libraries are imported as needed by the clients of libraries that rely on them, unlike features implemented by language-external means, avoiding the compatibility problem.
+We argue that, due to these problems, taking a language-external approach to realizing a new feature should be considered harmful and avoided whenever possible.
+The goal of the research being proposed here is to design \emph{language-integrated extension mechanisms} that give developers the ability to define in libraries new edit-time and compile-time features of the system that have previously been defined centrally\footnote{One might compare today's monolithic programming systems to  {centrally-planned} economies, whereas extensible\- systems more closely resemble modern market economies.} so that such language-external approaches are less frequently necessary.
+More specifically, we will show how control over aspects of \textbf{syntax}, \textbf{typechecking}, \textbf{elaboration} and \textbf{editor services} can be delegated to user-defined logic distributed in {libraries}, as illustrated in Figure \ref{approaches}b.
+Such libraries are called \emph{active libraries} because, rather than being passive clients of features already available in the system, they contain logic invoked by the system during development or compilation to provide new features  \cite{activelibraries}.
+Features implemented within active libraries can be imported as needed by the clients of libraries that rely on them, unlike features implemented by language-external means, seemingly avoiding the client compatibility problem.
 
-We must proceed with caution, however: critical issues having to do with {safety} must be overcome before library-based extension mechanisms can be introduced into a programming system, because if too much control over such core aspects of the system is given  to developers, the system may become unreliable.
+We must proceed with caution, however: critical issues having to do with {safety} must be overcome before language-integrated extension mechanisms can be introduced into a programming system, because if too much control over such core aspects of the system is given  to developers, the system may become unreliable.
 %For example, an extension could weaken important metatheoretic guarantees previously provided by the system. 
 Type safety, for example, may not hold if the static and dynamic semantics of the language can be modified or extended arbitrarily from within libraries.
-Furthermore, even if extensions can be shown not to cause such problems in isolation, there may still be conflicts between extensions that could weaken their semantics and lead to subtle problems at link-time.
-For example, if two active libraries defined differing semantics for the same syntactic form, the issue would only manifest itself when both libraries were imported somewhere within the same program.
+Furthermore, even if extensions can be shown not to cause such problems in isolation, there may still be conflicts between extensions that could weaken their semantics, leading to subtle problems that only appear at link-time.
+For example, if two active libraries introduce the same syntactic form but back it with differing (valid) semantics, the issue would only manifest itself when both libraries were imported somewhere within the same compilation unit.
 These kinds of safety issues have plagued previous attempts to design language-integrated extensibility mechanisms.% To prevent them, our mechanisms will organize  extension logic around types to guarantee that extensions are both safe in isolation and also safely composable in any combination.

### 2014-01-14 thesis-proposal/problem.tex
Done sections 1 and 2 of proposal

@@
 In some cases, desirable features must be omitted entirely due to concerns about  interoperability.
-F\#, for example, is based on Ocaml, but due to the need for bidirectional interoperability with CLI languages, it cannot support features like polymorphic variants, modules or functors \cite{ocaml-manual}.
+F\#, for example, is based on Ocaml, but due to the need for bidirectional interoperability with CLI languages, it cannot support features like polymorphic variants, modules or functors \cite{ocaml-manual} because they have no straightforward analogs in the object-oriented intermediate language of the CLI.
 %\end{itemize}

### 2014-01-16 thesis-proposal/acc.tex
Updated beginning of proposal
continuing work on proposal
Full rough draft.

@@
 % !TEX root = omar-thesis-proposal.tex
-\section{Active Code Completion}
+\section{Active Code Completion}\label{acc}
+Software developers today make heavy use of the code completion support found in modern source code editors  \cite{murphy_how_2006}.
+Most editors provide code completion in the form of a floating menu containing  contextually-relevant variables, fields, methods, types and other code snippets.
+By navigating and selecting from this menu, developers are able to avoid many common spelling and logic errors, eliminate unnecessary keystrokes and explore unfamiliar APIs without incurring the mental overhead associated with switching to an external documentation tool or API browser.
+% The way code completion menus brought documentation systems into the editor, active code completion interfaces bring external tools into the editor.
+
+Several refinements and additions to the code completion menu have previously been suggested in the literature.
+These have focused on leveraging additional sources of information, such as databases of usage history \cite{robbes_how_2008}\cite{HouPletcher2011}, inheritance information \cite{HouPletcher2011}, API-specific information \cite{HouPletcher2011}\cite{Lee+2008}, partial abbreviations \cite{Han+2009}, examples extracted from code repositories \cite{bruch_learning_2009}\cite{Brandt+2010} and crowdsourced information \cite{mooty_calcite:_2010}\cite{SnipMatch}, to increase the relevance and sophistication of the featured menu items.
+As with the standard form of code completion, many of these sources of data can also be utilized via external tools (e.g.
+Calcite \cite{mooty_calcite:_2010} uses information that could already be accessed using the Jadeite \cite{conf/vl/StylosFYM09} tool).
+Empirical evidence presented in these studies, however, suggests that directly integrating these kinds of tools into the editor is particularly effective.
+For example, users of the Calcite tool completed 40\% more tasks in a lab study (unfortunately, a Jadeite control group was not included.)
+
+In all of these systems, the code completion interface has remained primarily menu-based.
+When an item is selected, code is inserted immediately, without further input from the developer.
+These systems are also difficult to extend: a fixed strategy determines the completions that are available, so library providers cannot directly specify new domain-specific or contextually-relevant logic.
+In this paper we propose a technique called {\it active code completion} that eliminates these restrictions\footnote{Portions of this work previously appeared in a poster abstract \cite{ACC_VLHCC}.}.
+This  makes developing and integrating a broad array of highly-specialized developer tools directly into the editor, via the familiar code completion command, significantly simpler.
+This technique is motivated by the evidence discussed above and further evidence provided in this paper that developers prefer, and make more effective use of, tools that do not require leaving the immediate editing environment.
+
+In this paper, we discuss active code completion in the context of object construction. For example, consider the following Java code fragment:
+
+\begin{lstlisting}
+  public Color getDefaultColor() \{
+      return \textvisiblespace
+\end{lstlisting}
+
+\begin{figure*}\label{color}
+\begin{center}
+\includegraphics[width=40pc]{color_palette.png}\end{center}
+\caption{(a) An example code completion palette associated with the \texttt{Color} class. (b) The source code generated by this palette.}
+\end{figure*}
+
+If the developer invokes the code completion command at the indicated cursor position (\textvisiblespace), the editor  looks  for a {\it palette definition} associated with the {\it type} of the expression being entered, which in this case is  \verb|Color|.
+If an associated palette is found, a menu item briefly describing this palette is added to the standard code completion menu.
+When selected, the corresponding palette is shown, replacing the standard code completion menu.
+Figure 1a gives an example of a simple palette that may be associated with the \verb|Color| class\footnote{A video demonstrating this process is available at \\\url{http://www.cs.cmu.edu/~NatProg/graphite.html}.}.
+
+The developer can interact with such palettes to provide parameters and other information related to her intent, and receive immediate feedback about the effect these choices will have on the behavior of the object being constructed.
+When this interaction is complete, the palette generates appropriate source code for insertion at the cursor.
+Figure 1b shows the inserted code after the user presses \textsc{enter}.
+
+In accordance with best practices, we sought to address the following questions before designing and implementing our active code completion system:
+
+\begin{itemize}
+\item What {\it specific} use cases exist for this form of active code completion in a professional development setting? 
+\item What {\it general} criteria are common to types that would and would not benefit from an associated palette?
+\item What are some relevant usability and design criteria for palettes designed to address such use cases?
+\item What capabilities must the underlying active code completion system provide to enable these use cases and user interface designs?
+\end{itemize}
+
+To help us answer these questions, we conducted a survey of 473 professional developers (Section II).
+Their responses, along with information gathered from informal interviews and code corpus analyses, revealed a number of non-trivial functional requirements for palette interfaces as well as the underlying active code completion architecture (Section III).
+Participants also suggested a large number of use cases, demonstrating the broad applicability of this technique.
+We organize these into several broad categories (Section IV).
+
+Next, we describe Graphite, an Eclipse plug-in that implements the active code completion architecture for the Java programming language (Section V), allowing Java library developers to associate custom palettes with their own classes.
+We describe several design choices that we made to satisfy the requirements discovered in our preliminary investigations and briefly examine necessary trade-offs.
+
+Finally, we conducted a pilot lab study with a more complex palette, implemented using Graphite, that assists developers as they write regular expressions (Section VI).
+The study provides specific evidence in support of the broader claim that highly-specialized tools that are integrated directly with the editing environment are particularly useful.
+We conclude that active code completion systems like Graphite are useful because they make developing, deploying and discovering such tools fundamentally simpler.
+
+\subsection{Remaining Tasks}
+\begin{itemize}
+\item Related work: Murphy-Hill and Murphy, CSCW paper on peer interaction to enable discovery of tools; 
+Improving program navigation with an active help system. CASCON 2010.; 
+Improving software engineer's fluency by recommending dev environment commands. FSE 2012.
+\item Discuss safety more specifically.
+\end{itemize}

### 2014-01-16 thesis-proposal/parsing.tex
updated proposal
Updated beginning of proposal
continuing work on proposal
Full rough draft.

--- 2013-09-10 thesis-proposal/acc.tex
@@
 % !TEX root = omar-thesis-proposal.tex
-\section{Active Code Completion}
+\lstset{tabsize=2, 
+basicstyle=\ttfamily\fontsize{8pt}{1em}\selectfont, 
+commentstyle=\itshape\rmfamily, 
+stringstyle=\ttfamily,
+numbers=left, numberstyle=\scriptsize\color{gray}\ttfamily, language=ML,moredelim=[il][\sffamily]{?},mathescape=true,showspaces=false,showstringspaces=false,xleftmargin=15pt,escapechar=@, morekeywords=[1]{let,fn,val,def,casetype,objtype,metadata,of,*,datatype,new,valAST},deletekeywords={for,double,in},classoffset=0,belowskip=\smallskipamount,
+moredelim=**[is][\color{cyan}]{SSTR}{ESTR},
+moredelim=**[is][\color{OliveGreen}]{SHTML}{EHTML},
+moredelim=**[is][\color{purple}]{SCSS}{ECSS},
+moredelim=**[is][\color{brown}]{SSQL}{ESQL},
+moredelim=**[is][\color{orange}]{SCOLOR}{ECOLOR},
+moredelim=**[is][\color{magenta}]{SPCT}{EPCT}, 
+moredelim=**[is][\color{gray}]{SNAT}{ENDNAT}, 
+moredelim=**[is][\color{blue}]{SURL}{EURL},
+moredelim=**[is][\color{SeaGreen}]{SQT}{EQT},
+moredelim=**[is][\color{Periwinkle}]{SGRM}{EGRM},
+moredelim=**[is][\color{YellowGreen}]{SID}{EID}
+}
+\lstloadlanguages{Java,VBScript,XML,HTML}
+\let\li\lstinline
+
+\section{Type-Specific Languages}\label{aparsing}
+By using a general-purpose abstraction mechanism to encode a data structure, one immediately benefits from a body of established reasoning principles and primitive operations.
+For example, inductive datatypes can be used to express data structures like lists: intuitively, a list can either be empty, or be broken down into a value (its \emph{head}) and another list (its \emph{tail}).
+In an ML-like language, this concept is conventionally written:
+\begin{lstlisting}[numbers=none]
+datatype 'a list = Nil | Cons of 'a * 'a list
+\end{lstlisting}
+By encoding lists in this way, we can reason about them by structural induction, construct them by choosing the appropriate case and inspect them by pattern matching.
+The programmer only needs to provide an encoding of the structure of lists; the semantics are filled in by the general-purpose abstraction mechanism.
+
+While inheriting semantics can be quite useful, inheriting associated general-purpose syntax can sometimes be a liability.
+For example, few would claim that writing a list of numbers as a sequence of \li{Cons} cells is convenient:
+\begin{lstlisting}[numbers=none]
+Cons(1, Cons(2, Cons(3, Cons(4, Nil))))
+\end{lstlisting}
+
+Because lists are a common data structure, many languages provide specialized notation for constructing them, e.g.
+\li{[1, 2, 3, 4]}.
+This notation is semantically equivalent to the general-purpose notation shown above, but brings cognitive benefits by drawing attention to the content of the list, rather than the nature of the encoding.
+More specifically, it is more \emph{terse}, \emph{visible} and \emph{maps more closely} to the intuitive notion of a list, to use terminology from the literature on the cognitive dimensions of notations \cite{green1996usability}.
+
+Although number, string and list literals are nearly ubiquitous features of modern languages, some languages also  provide specialized notation for other common data structures, like maps and sets, data formats, like XML and JSON, query languages, like regular expressions and SQL, and markup languages, like HTML.
+For example, a language with built-in syntax for HTML and SQL, with type-safe interpolation of host language terms via curly braces, might allow:
+\begin{lstlisting}
+let webpage : HTML = SHTML<html><body><h1>Results for {EHTMLkeywordSHTML}</h1>
+  <ul id="results">{EHTMLto_list_items(query(db, 
+    SSQLSELECT title, snippet FROM products WHERE {ESQLkeywordSSQL} in titleESQL)SHTML}
+  </ul></body></html>EHTML
+\end{lstlisting}
+to mean:
+\begin{lstlisting}
+let webpage : HTML = HTMLElement(Empty, [BodyElement(Empty,
+  [H1Element(Empty, [TextNode($\texttt{"}$SSTRResults for $\texttt{"}$ESTR + keyword)]), 
+  ULElement((add Empty ($\texttt{"}$SSTRid$\texttt{"}$ESTR, $\texttt{"}$SSTRresults$\texttt{"}$ESTR)), to_list_items(query(db, 
+    SelectStmt([$\texttt{"}$SSTRtitle$\texttt{"}$ESTR, $\texttt{"}$SSTRsnippet$\texttt{"}$ESTR], $\texttt{"}$SSTRproducts$\texttt{"}$ESTR, 
+      [WhereClause(InPredicate(StringLit(keyword), $\texttt{"}$SSTRtitle$\texttt{"}$ESTR))]))))]]]
+\end{lstlisting}
+
+When a specialized notation is not available, and equivalent general-purpose notation is too cognitively demanding for comfort, developers typically turn to run-time mechanisms to make constructing data structures more convenient.
+Among the most common strategies in these situations is to simply use a string representation that is parsed at run-time.
+Developers are frequently  tempted to write the example above as:
+\begin{lstlisting}
+let webpage : HTML = parse_html($\texttt{"}$SSTR<html><body><h1>Results for ESTR$\texttt{"}$+keyword+$\texttt{"}$SSTR</h1>
+  <ul id=\$\texttt{\color{cyan}"}$results\$\texttt{\color{cyan}"}$>$\texttt{"}$ESTR + to_string(to_list_items(query(db, parse_sql(
+  	$\texttt{"}$SSTRSELECT title, snippet FROM products WHERE '$\texttt{"+keyword+"}$' in title$\texttt{"}$ESTR)))) + 
+  $\texttt{"}$SSTR</ul></body></html>$\texttt{")}$
+\end{lstlisting}
+
+Though recovering much of the notational convenience of the literal version, it is still more awkward to write, requiring explicit conversions to and from structured representations and escaping when the syntax of the language clashes with the syntax of string literals (line 2).
+But code like this also causes a number of more serious problems beyond cognitive load.
+Because parsing occurs at run-time, syntax errors will not be discovered statically, causing potential problems in production scenarios.
+Run-time parsing also incurs performance overhead, particularly relevant when code like this is executed often (as on a heavily-trafficked website, or in a loop).
+But the most serious issue with this code is that it is fundamentally insecure: it is vulnerable to cross-site scripting attacks (line 1) and SQL injection attacks (line 3).
+For example, if a user provided the keyword \li{'; DROP TABLE products --}, the entire product database could be erased.
+These attack vectors are considered to be two of the most serious security threats on the web today \cite{owasp2013}.
+Although developers are cautioned to sanitize their input, it can be difficult to verify that this was done correctly throughout a codebase.
+The most straightforward way to avoid these problems is to use structured representations throughout the codebase, aided by specialized notation like that above \cite{Bravenboer:2007:PIA:1289971.1289975}.
+
+To emphasize that this is a common problem, let us return to considering regular expression literals.
+It is quite tedious to write out a regular expression in a structured manner.
+A simple regular expression like \verb!(\d\d):(\d\d)\w*((am)|(pm))!
+representing times might be written:
+\begin{lstlisting}
+Seq(Group(Seq(Digit, Digit), Seq(Char($\texttt{"}$SSTR:ESTR$\texttt{"}$), Seq(Group(Seq(Digit, Digit)), 
+  Seq(ZeroOrMore(Whitespace), Group(Or(Group(Seq(Char($\texttt{"}$SSTRaESTR$\texttt{"}$), Char($\texttt{"}$SSTRmESTR$\texttt{"}$))), 
+  Group(Seq(Char($\texttt{"}$SSTRpESTR$\texttt{"}$), Char($\texttt{"}$SSTRmESTR$\texttt{"}$))))))))))
+\end{lstlisting}
+This is clearly more cognitively demanding, both when authoring the regular expression and when reading it.
+Among the most common strategies in these situations, for users of both object-oriented and functional languages, is to simply use a string representation that is parsed at run-time.
+\begin{lstlisting}
+rx_from_str($\texttt{"}$SSTR(\\d\\d):(\\d\\d)\\w*((am)|(pm))ESTR$\texttt{"}$)
+\end{lstlisting}
+%%
+%%For example, in languages without SQL literals, developers can implement a builder pattern:
+%%\begin{lstlisting}
+%%new SQLQuery().SELECT("*").FROM("table").WHERE("username").Eq(username)
+%%\end{lstlisting}
+This is problematic, for the same reasons as described above.
+
+As we will examine further in our corpus analysis, situations like this, where specialized notation is  necessary to maintain strong correctness, performance and security guarantees while avoiding unacceptable cognitive overhead, are quite common.
+Today, implementing new notations within an existing language requires the cooperation of the language designer.
+A primary reason for this is that, with conventional parsing strategies, not all notations can safely coexist, so a designer is needed to make choices about which syntactic forms are available and what their semantics are.
+For example, conventional notations for sets and maps are both delimited by curly braces.
+When Python introduced set literals, it chose to distinguish them based on whether the literal contained only values, or key-value pairs.
+But this causes an ambiguity with the syntactic form \verb|{ }| -- should it mean an empty set or an empty map (called a dictionary in Python)?
+The designers of Python chose the latter interpretation (for backwards compatibility reasons, in this case).
+
+Languages that allow users to introduce new syntax from within libraries hold promise, but because there is no longer a designer making decisions about such ambiguities, the burden of resolving them falls to the clients of extensions.
+For example, SugarJ \cite{erdweg2011sugarj} and other extensible languages generated by Sugar* \cite{erdweg2013framework} allow providers to extend the base syntax of the host language (e.g.
+Java) with new forms, like set and map literals.
+New forms are imported transitively throughout a program.
+To resolve syntactic ambiguities that arise, clients must manually augment the composed grammar with new rules that allow them to choose the correct interpretation explicitly.
+This is both difficult to do, requiring an understanding of the underlying parser technology (in Sugar*, GLR parsing using SDF) and increases the cognitive load of using the conflicting notations (e.g.
+both sets and dictionaries) in the same file.
+These kinds of conflicts occur in a variety of circumstances: HTML and XML, different variants of SQL, JSON literals and dictionaries, or simply different implementations (``desugarings'') of the same specialized syntax (e.g.
+two regular expression engines) can all cause problems.
+
+In this work, we will describe an alternative parsing strategy that avoids these problems by shifting responsibility for parsing certain \emph{generic literal forms} into the typechecker.
+The typechecker, in turn, defers responsibility to user-defined types, by treating the body of the literal as a term of the   \emph{type-specific language (TSL)} associated with the type it is being checked against.
+The TSL is responsible for rewriting this term to ultimately use only general-purpose notation.
+This strategy avoids the problem of conflicting syntax, because neither the base language nor TSLs are ever extended directly.
+It also avoids semantic conflicts -- the meaning of a form like \verb|{ }| can differ depending  on its type, so it is safe to use it for empty sets, dictionaries and other data structures, like JSON literals.
+This frees these common notations from the variant of a  data structure built into the standard library, which sometimes does not provide the exact semantics that a programmer needs (for example, Python dictionaries do not preserve order, while JSON does).
+%\begin{lstlisting}
+%let empty_set : Set = { }
+%let empty_dict : Dict = { }
+%let empty_json : JSON = { }
+%\end{lstlisting}
+\subsection{Wyvern}
+We develop our work as a variant of an emerging programming language being developed by our group called Wyvern \cite{Nistor:2013:WST:2489828.2489830}.
+To allow us to focus on the essence of our proposal, the variant of Wyvern we will describe in this thesis is simpler than the variant previously described: it is purely functional (there are no effects or mutable state) and it does not enforce a uniform access principle for objects (fields can be accessed directly).
+We also add recursive sum types, which we call \emph{case types}, that operate similarly to datatypes in ML.
+One can refer to the version of the language described in this thesis as \emph{TSL Wyvern} when the variant being discussed is not clear.
+
+We begin with an example in Fig.
+\ref{f-example} showing several different TSLs being used to define a fragment of a web application showing search results from a database.
+We will review this example below to develop intuitions about TSLs in Wyvern.
+\todo{describe colors}
+\begin{figure}[t]
+\begin{lstlisting}
+let imageBase : URL = <SURLimages.example.comEURL>
+let bgImage : URL = <SURL%EURLimageBaseSURL%/background.pngEURL>
+new : SearchServer
+  def resultsFor(searchQuery : String, page : Nat) : Unit =
+    serve(~) (* serve : HTML -> Unit *)
+SHTML      :html
+        :head
+          :title Search Results
+          :style EHTML~
+SCSS            body { background-image: url(%ECSSbgImageSCSS%) }
+            #search { background-color: %ECSS`SCOLOR#aabbccECOLOR`.darken(SPCT20pctEPCT)SCSS% }
+ECSSSHTML        :body
+          :h1 Results for {EHTMLsearchQuerySHTML}
+          :div[id="search"]
+            Search again: {EHTMLSearchBox($\texttt{"}$SSTRGo!ESTR$\texttt{"}$)SHTML}
+          {EHTML (* fmt_results : (DB, SQLQuery, Nat, Nat) -> HTML *)
+            fmt_results(db, ~, SNAT10ENDNAT, page)
+              SSQLSELECT * FROM products WHERE {ESQLsearchQuerySSQL} in titleESQL
+          SHTML}
+\end{lstlisting}
+\vspace{-8px}
+\caption{Wyvern Example with Multiple TSLs}
+\label{f-example}
+\vspace{-10px}
+\end{figure}
+\subsection{Inline Literals}
+\begin{figure}[t]
+\begin{lstlisting}
+<SURLliteral body here, <inner angle brackets> must be balancedEURL>
+{SURLliteral body here, {inner braces} must be balancedEURL}
+[SURLliteral body here, [inner brackets] must be balancedEURL]
+`SURLliteral body here, ``inner backticks`` must be doubledEURL`
+$\texttt{'}$SURLliteral body here, ''inner single quotes'' must be doubledEURL$\texttt{'}$
+$\texttt{"}$SURLliteral body here, ""inner double quotes"" must be doubledEURL$\texttt{"}$
+SURL12xyzEURL (* no delimiters necessary for number literals *)
+\end{lstlisting}
+\vspace{-8px}
+\caption{Inline Generic Literal Forms}
+\vspace{-10px}
+\label{f-delims}
+\end{figure}
+Our first TSL appears on line 1.
+The type of \li{imageBase} is \li{URL}, a \emph{structural type} containing several fields representing the components of a URL: its protocol, domain name, port, path and so on.
+We could create a value of type \li{URL} using general-purpose notation:
+\begin{lstlisting}
+let imageBase : URL = new
+  val protocol : String = $\text{"}$SSTRhttpESTR$\texttt{"}$
+  val subdomain : String = $\texttt{"}$SSTRimagesESTR$\texttt{"}$
+  val domain : String = $\texttt{"}$SSTRexampleESTR$\texttt{"}$
+  (* ... *)
+\end{lstlisting}
+This is tedious.
+If the \li{URL} type has a TSL associated with it, we can instead instantiate precisely this value using conventional notation for URLs by placing it in a \emph{generic literal}, \li{<SURLimages.example.comEURL>}.
+The type annotation on \li{imageBase} implies that this literal's \emph{expected type} is \li{URL}, so the \emph{body} of the literal (the characters between the angle brackets, in blue) is governed by the \li{URL} TSL.
+This TSL will parse the body {at compile-time} to produce a Wyvern AST that explicitly instantiates a new object of type \li{URL}  (see below).
+Any other delimited form in Fig.
+\ref{f-delims} could equivalently be used if the constraints shown are obeyed.
+
+In addition to supporting conventional notation for URLs, this TSL supports \emph{typed interpolation}
+of a \li{URL} expression to form a larger URL. The interpolated term is delimited by percent signs,
+as seen on line 2 of Fig. \ref{f-example}. The TSL parses code between percent signs  as a Wyvern expression with expected type \li{URL}. The TSL then assumes the interpolated expression is of this
+type to construct an AST. Untyped---and thus unsafe---interpolation of strings
+does not occur.
+Note that the delimiters that can be used to go from Wyvern to a TSL are determined by Wyvern (those in Fig.
+\ref{f-delims}) while the delimiters that can be used to go from a TSL back to Wyvern are chosen by the TSL (see Sec.
+\ref{ss:implementing-interpolation}).
+
+%\subsection{General-Purpose Notation for Object and Case Types}
+%The general-purpose introductory form for object types is \li{new}.
+This form is a syntactic \emph{forward reference} to the layout-delimited block of {definitions} beginning on the line immediately after the line \li{new} appears in (line 4 in this case), and ending when the indentation level has returned to the baseline, or when the file ends (after line 19 in this case).
+An object in TSL Wyvern can contain methods, introduced using \li{def}, and fields, introduced using \li{val}.
+Here we have just a single method, \li{serve_results} taking two arguments.
+Object types in TSL Wyvern are simple structural interfaces that constrain the signatures of fields and methods.
+The \emph{type ascription} around \li{new} checks that the object being introduced satisfies the signature of \li{SearchServer} (not shown).\todo{discuss case types}
+\subsection{Layout-Delimited Literals}
+On line 5 of Fig.
+\ref{f-example}, we see a call to a function \li{serve}, not shown, which has type \li{HTML -> Unit}.
+Here, \li{HTML} is a user-defined \emph{case type}, having cases for each HTML tag as well as some other structures.
+Declarations  of some of these cases can be seen on lines 2-6 of Fig.
+\ref{f-htmltype} (TSL Wyvern also includes simple product types for convenience, written \li{T1 * T2}).
+The general-purpose introductory form for a case type like \li{HTML} is, e.g., \li{HTML.BodyElement((attrs, child))}.
+But, as discussed above, using this syntax can be cognitively demanding.
+Thus, we associate a TSL with \li{HTML} that provides a simplified notation for writing HTML, shown being used on lines 6-20 of Fig.
+\ref{f-example}.
+This literal body is layout-delimited, rather than delimited by explicit tokens as in Fig.
+\ref{f-delims}, and introduced by a form of \emph{forward reference}, written \li{~} (``tilde''), on the previous line.
+Because the forward reference occurs in a position where the expected type is \li{HTML}, the literal body is governed by that type's TSL.
+The forward reference will be replaced by a general-purpose term of type \li{HTML} generated from the literal body by the TSL during compilation.
+\begin{figure}[t]
+\begin{lstlisting}[escapechar=$]
+casetype HTML 
+    Empty of Unit
+  | Text of String
+  | Seq of HTML * HTML 
+  | BodyElement of Attributes * HTML
+  | StyleElement  of Attributes * CSS
+  | ...
+  metadata = new : HasParser
+    val parser : Parser = ~
+SGRM      start -> ':body'= child::start>
+        EGRM`SQTHTML.BodyElement(([], %EQTchildSQT%))EQT`
+SGRM      start -> ':style'= e::EXP[NEWLINE]>
+        EGRMlet empty_attrs : Attributes = []
+        Exp.CaseIntro(Type.Var($\texttt{"}$SIDHTMLEID$\texttt{"}$), $\texttt{"}$SIDStyleElementEID$\texttt{"}$,
+          Exp.ProdIntro(valAST(empty_attr), e))
+SGRM      start -> '{'= e::EXP['}']>EGRM
+        `SQT%EQTeSQT% : HTMLEQT`
+\end{lstlisting}
+\vspace{-8px}
+\caption{A Wyvern case type with an associated TSL.}
+\vspace{-10px}
+\label{f-htmltype}
+\end{figure}
+
+\subsection{Implementing a TSL}
+Portions of the implementation of the TSL for \li{HTML} are shown on lines 8-17 of Fig.
+\ref{f-htmltype}, written as a declarative grammar.
+A \li{Parser} is associated with a type using \li{metadata}, as shown.
+The \li{metadata} of a type \li{t} is simply a value that is associated with the type at both compile-time and run-time.
+It can be accessed using the syntactic form \li{t.metadata}.
+In this case, it is an object with a single field, called \li{parser} of type \li{Parser} (defined in Wyvern's standard library).
+A type equipped with a parser in this way is an example of an \emph{active type}.
+
+A grammar can be thought of as specialized notation that generates a parser.
+Since we have a type for parsers, \li{Parser}, we can implement a grammar-based parser generator  itself  as a TSL.
+In this case, we are basing our grammar formalism on the layout-sensitive formalism developed by Adams \cite{Adams:2013:PPI:2429069.2429129} -- Wyvern is itself layout-sensitive and has a grammar that can be written down using this formalism, so it makes sense to support it for layout-sensitive TSLs as well.
+Most aspects of this formalism are completely conventional.
+Each non-terminal (e.g.
+\li{start}) is defined by a number of disjunctive productions, each introduced using \li{->}.
+Each production defines a sequence of terminals (e.g.
+\li{':body'}) and non-terminals (e.g.
+\li{start}), which can have names (e.g \li{child}) associated with them using \li{::}.
+Unique to Adams' formalism is that each terminal and non-terminal in a production can also have a \emph{layout constraint} following it.
+Here, the layout constraints can be either \li{=} (meaning that the leftmost column of the annotated term must be aligned with that of the parent term), \li{>} (the leftmost column must be indented further), \li{>=} (the leftmost column \emph{may} be indented further).
+Layout constraints are optional in TSLs.
+We will discuss this further when we formally describe Wyvern's layout-sensitive concrete syntax.
+
+Each rule is followed by a Wyvern expression, in an indented block, that implements the rewriting logic for the associated rule.
+This expression should have type \li|Exp|, a case type built into the standard library representing Wyvern expression ASTs.
+The ASTs generated by named non-terminals in the rule (e.g.
+\li{child}) are bound to the corresponding variables within this logic.
+Here, we show how to generate an AST using general-purpose notation for \li{Exp} (lines 13-15) as well as the more natural \emph{quasiquote} style (lines 11 and 18).
+Quasiquotes are simply the TSL associated with \li{Exp} and support the full Wyvern concrete syntax as well as an additional delimiter form, written with \li{%}s, that allows ``unquoting'': interpolating another AST into the one being generated.
+Again, interpolation is typesafe, structured and occurs at compile time.
+
+\subsection{Implementing Interpolation}\label{ss:implementing-interpolation}
+We have now seen several examples of interpolation. Within the TSL for \li{HTML}, for example, we see it used in several ways:
+
+\subsubsection{HTML Interpolation} At any point where a tag should appear, we can also interpolate a Wyvern expression of type \li{HTML} by enclosing it within curly braces (e.g.
+on line 13, 15 and 16-19 of Fig.
+\ref{f-example}).
+This is implemented on lines 17 and 18 of Fig.
+\ref{f-htmltype}.
+The special non-terminal \li{EXP[token]} signals a switch into the Wyvern grammar.
+The tokenstream will be parsed as a Wyvern expression until the token \li{token} is encountered \emph{where it would otherwise trigger a parse error}.
+In other words, the Wyvern grammar binds more tightly to itself than to any surrounding TSL, avoiding potential ambiguities.
+The AST for the parsed Wyvern expression is given an expected type, \li{HTML}, by simply surrounding it with a type ascription (line 18).
+Because interpolation must be structured (a string cannot be interpolated directly), injection and cross-site scripting attacks cannot occur.
+Safe interpolation of expressions of type \li{String} (where any inner angle brackets and other special characters are turned into HTML entities, e.g.
+\li{&lt;} for \li{<}) could similarly be implemented using another delimiter.
+
+\subsubsection{CSS Interpolation} After the \li{:style} tag appears (e.g.
+on line 9 of Fig.
+\ref{f-example}), instead of hard-coding CSS syntax into the HTML DSL, we instead wish to use the TSL associated with a type representing a CSS stylesheet: \li{CSS}.
+We do this by again interpolating a Wyvern expression (lines 12-15 of Fig.
+\ref{f-htmltype}), making sure that it appears in a position where the expected type is \li{CSS} (the second piece of data associated with the \li{StyleElement} constructor, in this case).
+Wyvern is given control until a full expression has been read and an unexpected newline appears (that is, a newline that does not introduce a layout-delimited block).
+Here we see a use of a layout-delimited TSL within another layout-delimited TSL.
+
+\subsubsection{Interpolation within the CSS TSL} The TSL for \li{CSS} itself has support for interpolation in a similar manner, choosing \li{%} as the delimiter.
+It chooses the type based on the semantics of the surrounding CSS form.
+For example, when a Wyvern expression appears inside \li{url}, as on line 10 of Fig.
+\ref{f-example}, it must be of type \li{URL}.
+When a Wyvern expression appears where a color is needed, the \li{Color} type is used.
+This type itself has a TSL associated with it that interprets CSS color strings, showing again that TSLs can be used within TSLs by simply escaping out to Wyvern, the host language, and then back in.
+In this case, we emphasize that TSLs produced structured values by calling the \li{darken} method on it to produce a new color.
+This method itself takes a \li{Percentage} as an argument.
+The TSL for this type accepts literal bodies containing numbers followed by \li{pct}, or simply a real number without a suffix.
+Numeric literals, because they begin with a number (and no other form in Wyvern can), do not require delimiters (Fig.
+\ref{f-delims}).
+
+\subsubsection{Interpolation within the SQLQuery TSL} The TSL used for SQL queries on line 18 of Fig.
+\ref{f-example} follows an identical pattern, allowing strings to be interpolated into portions of a query in a safe manner.
+This prevents SQL injection attacks from occurring while maintaining standard SQL syntax.
+
+\subsection{Formalization}
+A formal and more detailed description can be found in our paper draft\footnote{\url{https://github.com/wyvernlang/docs/tree/master/ecoop14}}.
+In particular\todo{should I include some/all of this in the proposal?}:
+\begin{enumerate}
+\item We provide a more complete layout-sensitive concrete syntax.
+We show how it can be written without the need for a context-sensitive lexer or parser and give a full specification for the layout-delimited literal form introduced by a forward reference, \li{~}, as well as other forms of forward-referenced blocks.
+\item We detail the general mechanism for associating metadata with a type.
+A TSL is then implemented by associating a parser (of type \li{Parser}) with a type.
+The parser is responsible for rewriting tokenstreams (of type \li{Tokenstream}) into Wyvern ASTs (of type \li{Exp}).
+These types are defined in the standard library.
+%\item This lower-level mechanism is general, but writing a hand-written parser and manipulating syntax trees manually is cognitively demanding.
+We observe that \emph{grammars} and \emph{quasiquotes} can both be seen as TSLs for parsers and ASTs respectively and discuss how to implement them as such.
+\item A na\"ive rewriting strategy would be \emph{unhygienic} -- it could allow for the inadvertent capture of local variables.
+We show a novel mechanism that ensures hygiene by requiring that the generated AST is closed except for subtrees derived from portions of the user's tokenstream that are interpreted as nested Wyvern expressions.
+We also show how to explicitly refer to local values available in the parser definition (e.g.
+helper functions) in a safe way.
+\item We formalize the static semantics and literal parsing rules of TSL Wyvern as a bidirectional type system.
+By distinguishing locations where an expression synthesizes a type from locations where an expression is being analyzed against a previously synthesized type, we can precisely state where generic literals can appear.
+This system also provides a formal specification of our hygiene mechanism.
+\item We provide several examples of TSLs throughout the paper, but to examine how broadly applicable the technique is, we conduct a simple corpus analysis, finding that string languages are used ubiquitously in existing Java code  (collaborative work with Darya Kurilova).
+\end{enumerate}
+
+\subsection{Remaining Tasks}
+The following tasks remain to be completed:
+\begin{enumerate}
+\item We must write down the full formal semantics (including rules that we omitted for concision in the paper draft) in a technical report, as well as provide more rigorous proofs of the metatheory.
+\item We must further consider aspects of hygiene.
+In particular, we do not have a clean mechanism for preventing unintentional variable \emph{shadowing}, only unintentional variable \emph{capture}.
+It may be possible to prevent shadowing by making it impossible for variables to be introduced into interpolated Wyvern expressions (so that function application is the only way to pass data from a TSL to Wyvern code).
+\item The corpus analysis we conducted was preliminary. We must perform this in a more complete and rigorous manner (collaborative work with Darya Kurilova).
+%\item The current implementation of Wyvern does not include all of these mechanisms as described.
+Although the implementation of Wyvern is not a project I am responsible for, I plan on working with the student who is doing this to implement as much of what we have described as possible.\todo{should I drop this from the proposal?
+I don't want to promise that everything will be implemented because that is largely contingent on Benjamin's effort, not mine.}
+\end{enumerate}
+%
+%\subsection{TODO}
+%- Figure out separate compilation (e.g. of CSS)

### 2014-01-16 thesis-proposal/problem.tex
Full rough draft.

@@
 Due to the increasing diversity and complexity of modern problem domains and hardware platforms, it has become clear that no small  collection of general-purpose constructs backed by a simple, all-purpose implementation can fully satisfy these criteria in all cases.
-Instead,  researchers and domain experts (collectively, \emph{providers}) continue to develop  specialized notations,  static semantics, dynamic semantics, implementation strategies, optimizations, run-time systems and tools (collectively, \emph{features})  designed to address these modern challenges in different circumstances.
+Instead,  researchers and domain experts (collectively, \emph{providers}) continue to develop  specialized notations,  static and dynamic semantics, implementation strategies, optimizations, run-time systems and tools (collectively, \emph{features})  designed to address these modern challenges in different circumstances.
 %, researchers and domain experts must continue to design and develop, in a principled manner, specialized notations, type systems, optimizations, run-time systems and tools (collectively, \emph{features}) to support developers as they work in increasingly complex problem domains and on modern hardware systems.
@@
 Unfortunately this is often infeasible because libraries are vehicles for specifying a program in terms of a monolithic collection of existing features.
-From the perspective of a library, the language's syntax, type system and semantics are fixed in advance, the compiler and run-time system are ``black box'' implementations of this fixed language, and the other tools, like code editors and debuggers, operate according to domain-agnostic protocols also based only on a fixed language specification.
+From the perspective of a library, the language's syntax, type system and dynamic semantics are fixed in advance, the compiler and run-time system are ``black box'' implementations of this fixed language, and the other tools, like code editors and debuggers, operate according to domain-agnostic protocols also based only on a fixed language specification.
 As a result, providers of new system features must often take \emph{language-external approaches}, sometimes designing a new system altogether.
 We will argue that such approaches are problematic, and that taking them has led to an unnecessary gap between research and practice.
-In place of these approaches, we will advocate for \emph{language-integrated extensibility mechanisms}, and show how, by organizing new features around {types} enforcing certain constraints, these mechanisms can be made both safe and highly expressive.
+In place of these approaches, we will advocate for \emph{language-integrated extensibility mechanisms}, and show how, by organizing new features around {types} and enforcing certain constraints, these mechanisms can be made both safe and highly expressive.
 We call a type that introduces new features an \emph{active type} and programming systems organized around active types \emph{actively-typed programming systems}.
@@
 When a type error is found, an intelligible error message is provided.
-\item \textbf{Elaboration logic} that partially or fully compiles known regular expressions into the efficient internal representation that will be used by the regular expression matching engine (e.g., a finite automata \cite{Thompson:1968:PTR:363347.363387}) ahead of time.
+\item \textbf{Translation logic} that partially or fully compiles known regular expressions into the efficient internal representation that will be used by the regular expression matching engine (e.g., a finite automata \cite{Thompson:1968:PTR:363347.363387}) ahead of time.
 In most languages, this compilation step occurs at run-time, even if the pattern is fully known at compile-time, thereby introducing performance overhead into programs.
@@
 Unfortunately, it is impossible to fully encode the syntax and the specialized static and dynamic semantics described above in terms of general-purpose notations and abstractions like objects or inductive datatypes.
-Library providers have thus needed to compromise, typically by asking clients to provide regular expressions as strings, thereby deferring parsing, typechecking and elaboration to run-time, which introduces performance overhead and can lead to unanticipated run-time errors (as shown in \cite{spishak2012type}) and security vulnerabilities (due to injection attacks when user inputs are spliced into patterns, for example).
+Library providers have thus needed to compromise, typically by asking clients to provide regular expressions as strings, thereby deferring parsing, typechecking and translation to run-time, which introduces performance overhead and can lead to unanticipated run-time errors (as shown in \cite{spishak2012type}) and security vulnerabilities (due to injection attacks when user inputs are spliced into patterns, for example).
 Similarly, useful tools for working with regular expressions are rarely integrated into editors, and even more rarely in a way that facilitates their discovery and use directly when the developer is manipulating regular expressions.
@@
 The goal of the research being proposed here is to design \emph{language-integrated extension mechanisms} that give developers the ability to define in libraries new edit-time and compile-time features of the system that have previously been defined centrally\footnote{One might compare today's monolithic programming systems to  {centrally-planned} economies, whereas extensible\- systems more closely resemble modern market economies.} so that such language-external approaches are less frequently necessary.
-More specifically, we will show how control over aspects of \textbf{syntax}, \textbf{typechecking}, \textbf{elaboration} and \textbf{editor services} can be delegated to user-defined logic distributed in {libraries}, as illustrated in Figure \ref{approaches}b.
+More specifically, we will show how control over aspects of \textbf{syntax}, \textbf{typechecking}, \textbf{translation} and \textbf{editor services} can be delegated to user-defined logic distributed in {libraries}, as illustrated in Figure \ref{approaches}b.
 Such libraries are called \emph{active libraries} because, rather than being passive clients of features already available in the system, they contain logic invoked by the system during development or compilation to provide new features  \cite{activelibraries}.

### 2014-01-16 thesis-proposal/timeline.tex
Full rough draft.

--- 2013-08-16 thesis-proposal/timeline.tex
@@
-\section{Research Plan}
-
-\subsection{Spring 2013}
-I will complete the work on Ace and active type theory and submit it for publication.
-
-\subsection{Summer 2013}
-I will complete the work on active code prediction and submit it for publication.
-
-\subsection{Fall 2013}
-I will finalize all work and write the final dissertation.
+% !TEX root = omar-thesis-proposal.tex
+\section{Timeline}
+\begin{itemize}
+\item @$\lambda$ - ICFP - Mar. 1
+\item Wyvern - ECOOP or OOPSLA - late March
+\item Ace - PLDI or OOPSLA - late March
+\item Then write everything up.
+\end{itemize}

### 2014-01-19 ace-pldi14/rebuttal.txt
Added reviews and rebuttal of PLDI14

We thank the reviewers for their detailed and insightful comments!

Due to the unusual number of reviews, we hope the committee does not mind that we exceeded the 500 word soft limit proportionally. To save time, the items most relevant to each reviewer are summarized below.

R1 - 1, 5, 6
R2 - 1, 5, 6, 9
R3 - 1, 2, 5, 6, 7
R4 - 2, 5, 6
R5 - 2, 3, 4, 5, 6, 8

Corrections:
1. R1 describes Ace as providing "syntax-directed" mechanisms, which would indeed be "routine". This is incorrect -- Ace uses a _type-directed_ mechanism to determine the semantics of an expression. That is, there is _not_ a direct mapping from syntax to semantics; the mapping depends on logic associated with the type of a designated subexpression. This represents an inversion of control relative to syntax-directed mechanisms. As R2 and R3 suggest by their characterizations of it, this mechanism is the main novel technical innovation Ace introduces.

2. R5 states that Ace's types are not "types in a conventional sense at all". This is incorrect -- Ace's types are constructs that statically classify expressions and are erased at run-time. There is no "dynamic typechecking" of Ace function bodies. Our theoretical work [1] connects our mechanism to type-level computation, further emphasizing that active types are clearly static types (this paper also contains a much simpler example, which R4 may be interested in). We will take care to emphasize this in the final version. Note to R3: the word "form" should be used for the syntactic form (e.g. Attribute); type is reserved for the static type.

3. R5 states that "Lisp-style macros are compositional" in the sense that we use the word. Figure 1 did in fact state this (though there are subtleties, see 5 below). Note  that macros do not permit non-local type system extensions.

4. R5 is incorrect that "all of the related work covered in Figure 1 is compositional". Systems like Qi/Shen, Typed Racket and Xroma do not prevent the introduction of conflicting rules that could cause problems by giving a type to constructs that another extension wished to classify as ill-typed or by violating previously maintained invariants (e.g. by violating the representational consistency check that Ace uses, see Sec. 3.1.3).

Clarifications:
5. We do agree with the reviewers that our explanation of extension composition in Ace could have been clearer, particularly because safe extensibility is a primary goal of Ace. There are two facets that we handle: "non-interference" -- each extension can claim exclusive control over a set of expressions, and "compositional reasoning" -- each extension can assume that expressions of another type are representationally consistent (have a consistent type in the target language, as determined by the trans_type method, cf. Sec. Sec. 3.1.3), making it possible to reason modularly about well-typedness of the translation. Both hinge on our novel type-directed dispatch protocol and our use of active types. We will take care to clarify this distinction. A fully-rigorous treatment of this topic requires a type theoretic formulation beyond the scope of this paper. Such a treatment can be found in [1].

6. Several reviewers (R1, R3, R5) had questions about binding and scope. We concur  that this could be discussed in more detail. As stated in Sec. 3.2, the active base controls local variable binding and sets up the typing context. The context can be structured as the base sees fit, but it is typically a dictionary mapping variable names to (active) types. Active types do not access it directly; recursing on a subexpression that contains variables will eventually delegate to the active base to extract type information.

6.1. Active types can add new contexts other than typing contexts to the "context" object by adding a new attribute to it upon first use of a construct that requires it. There is no guarantee that other types cannot manipulate this context, but if a globally unique attribute name is used, this will not occur accidentally (see 7 below) -- the new context will be statically threaded through the program unchanged until the next construct that intentionally accesses it is encountered. We hope to pursue a cleaner foundation for custom contexts in future work.

6.2. References to variables that are not local, like the names of other functions in scope, can also be handled by the active base by reference to the Python environment that the function is defined within. Ace captures this when a generic function is defined (we inadvertently did not state this). The special Python attribute ace_type designates an Ace type for an object in the Python environment. That type must determine how to lift this value from compile-time to run-time given the active target. This is how the OpenCL bindings in 2.6 work -- the ace_type attribute of OpenCL buffer objects (e.g. d_input) is an instance of clx.TypedBuffer parameterized by run-time type information available from the numpy object that was used to create the buffer (e.g. the numpy dtype attribute). R3 is correct that this interactive mode of use implies an "intermingling" of Python's dynamic semantics and Ace's static semantics -- this can be seen as a form of staged compilation (and we will describe it more clearly as such in Sec. 2.6).

6.3. Handling of nested function definitions (and lambdas) is also delegated entirely to the active base. For example, an active base supporting a C target could support a nested OpenCL-based function by using the OpenCL host API internally to compile it.

7. R3 asks for a description of a "well-behaved extension". These are extensions that do not use Python's metaprogramming facilities to modify other active types/bases or core Ace data structures directly, and that do not attempt to modify the base's typing context or contexts defined by other types. We will discuss this alongside our discussion of composability in the final version.

8. R5 states that LMS is the "system most similar" to Ace. Ace permits type system extensions while LMS uses Scala's monolithic (albeit expressive) type system directly. It is thus only comparable to the translation functionality Ace provides. We cited Delite but not LMS under "cross-compilation". We will cite both in the final version and discuss this distinction further in the text.

9. We are sympathetic to worries that the examples in Sec. 4 are not fully detailed. This is due to the 10-page limit. We will decrease the number of non-code examples to increase the level of detail for the ones with code in the final version. R2 suggests that we first develop more experience with Ace. We emphasize that we have developed a large number of extensions, including a full implementation of an industrial language -- OpenCL. Not discussed in this paper is a large distributed scientific simulation framework running on clusters of GPUs that has been used to conduct neurobiological circuit simulations written atop Ace (see Sec. XX of [2]). Several course and undergraduate research projects have now used Ace as well (e.g. [3]).

[1] TODO: ESOP draft
[2] TODO: SC draft
[3] TODO: Nathan's thesis

### 2014-01-21 ace-pldi14/rebuttal.txt
committed pldi14 rebuttal

@@
 We thank the reviewers for their detailed and insightful comments!
-
-Due to the unusual number of reviews, we hope the committee does not mind that we exceeded the 500 word soft limit proportionally.
-To save time, the items most relevant to each reviewer are summarized below.
+We would like to correct some errors that appear to have been made in the reviews and provide requested clarifications.
+Due to the unusual number of reviews, we hope the committee does not mind that we have exceeded the 500 word soft limit proportionally.
+The items most relevant to each reviewer are summarized below.
 
@@
 That is, there is _not_ a direct mapping from syntax to semantics; the mapping depends on logic associated with the type of a designated subexpression.
-This represents an inversion of control relative to syntax-directed mechanisms.
-As R2 and R3 suggest by their characterizations of it, this mechanism is the main novel technical innovation Ace introduces.
+This represents an inversion of control relative to conventional syntax-directed mechanisms.
+As R2 and R3 suggest by their characterizations of it, this is the main novel technical innovation Ace introduces.
 
@@
 There is no "dynamic typechecking" of Ace function bodies.
-Our theoretical work [1] connects our mechanism to type-level computation, further emphasizing that active types are clearly static types (this paper also contains a much simpler example, which R4 may be interested in).
+Our theoretical work [1] connects our mechanism to type-level computation, further emphasizing that active types are clearly static types.
+This paper also contains a much simpler example, as requested by R4.
 We will take care to emphasize this in the final version.
-Note to R3: the word "form" should be used for the syntactic form (e.g.
-Attribute); type is reserved for the static type.
+
+Note to R3: we use the word "form" for the syntactic form (e.g. Attribute, BinOp); "type" is reserved for static types.
 
@@
 R5 is incorrect that "all of the related work covered in Figure 1 is compositional".
-Systems like Qi/Shen, Typed Racket and Xroma do not prevent the introduction of conflicting rules that could cause problems by giving a type to constructs that another extension wished to classify as ill-typed or by violating previously maintained invariants (e.g.
+Systems like Qi/Shen, Typed Racket and Xroma allow extension writers to add to a "bag of rules" and so do not prevent the introduction of conflicting rules that could cause problems by, e.g., giving a type to constructs that another extension wished to classify as ill-typed or by violating previously maintained invariants (e.g.
 by violating the representational consistency check that Ace uses, see Sec.
@@
 Clarifications:
-5.
-We do agree with the reviewers that our explanation of extension composition in Ace could have been clearer, particularly because safe extensibility is a primary goal of Ace.
-There are two facets that we handle: "non-interference" -- each extension can claim exclusive control over a set of expressions, and "compositional reasoning" -- each extension can assume that expressions of another type are representationally consistent (have a consistent type in the target language, as determined by the trans_type method, cf.
+5. We do agree with the reviewers that our explanation of extension composition in Ace could have been clearer and made in one place in the paper. There are two facets that we handle: 
+  a) "non-interference" -- each extension has _exclusive_ control over a set of expressions, per the dispatch protocol
+  b) "compositional reasoning" -- each extension can assume that expressions of another type are representationally consistent (have a consistent type in the target language, as determined by the trans_type method, cf.
 Sec.
@@
 3.1.3), making it possible to reason modularly about well-typedness of the translation.
+
 Both hinge on our novel type-directed dispatch protocol and our use of active types.
-We will take care to clarify this distinction.
+We will take care to clarify this issue.
 A fully-rigorous treatment of this topic requires a type theoretic formulation beyond the scope of this paper.
@@
 Active types can add new contexts other than typing contexts to the "context" object by adding a new attribute to it upon first use of a construct that requires it.
-There is no guarantee that other types cannot manipulate this context, but if a globally unique attribute name is used, this will not occur accidentally (see 7 below) -- the new context will be statically threaded through the program unchanged until the next construct that intentionally accesses it is encountered.
+There is no guarantee that other types cannot manipulate this context, but if a globally unique attribute name is used, this will not occur accidentally (see 7 below); the new context will be statically threaded through the program unchanged until the next construct that intentionally accesses it is encountered.
 We hope to pursue a cleaner foundation for custom contexts in future work.
@@
 Ace captures this when a generic function is defined (we inadvertently did not state this).
-The special Python attribute ace_type designates an Ace type for an object in the Python environment.
+
+The special Python attribute ace_type can be used to designate an Ace type for an object in the Python environment.
 That type must determine how to lift this value from compile-time to run-time given the active target.
@@
 R3 asks for a description of a "well-behaved extension".
-These are extensions that do not use Python's metaprogramming facilities to modify other active types/bases or core Ace data structures directly, and that do not attempt to modify the base's typing context or contexts defined by other types.
+These are extensions that do not use Python's metaprogramming facilities to modify other active types/bases or core Ace data structures directly, and that do not attempt to modify the base's typing context or contexts defined by other extensions.
 We will discuss this alongside our discussion of composability in the final version.
@@
 R2 suggests that we first develop more experience with Ace.
-We emphasize that we have developed a large number of extensions, including a full implementation of an industrial language -- OpenCL.
-Not discussed in this paper is a large distributed scientific simulation framework running on clusters of GPUs that has been used to conduct neurobiological circuit simulations written atop Ace (see Sec.
-XX of [2]).
-Several course and undergraduate research projects have now used Ace as well (e.g.
+We emphasize that we have developed a large number of extensions, including a full implementation of an industrial language -- OpenCL -- using Ace.
+In addition, though not discussed in this paper, we have developed a large distributed scientific simulation framework running on clusters of GPUs that has been used to conduct neurobiological circuit simulations (see Sec.
+5 of [2]).
+Finally, several course and undergraduate research projects have now used Ace as well (e.g.
 [3]).
 
-[1] TODO: ESOP draft
-[2] TODO: SC draft
-[3] TODO: Nathan's thesis
+[1] https://github.com/cyrus-/papers/blob/master/esop14/att-esop14.pdf?raw=true
+[2] https://github.com/cyrus-/papers/blob/master/sc13/ace.pdf?raw=true
+[3] http://src.acm.org/2013/NathanFulton.pdf

### 2014-02-06 thesis-proposal/acc.tex
More progress on proposal

@@
 These systems are also difficult to extend: a fixed strategy determines the completions that are available, so library providers cannot directly specify new domain-specific or contextually-relevant logic.
-In this paper we propose a technique called {\it active code completion} that eliminates these restrictions\footnote{Portions of this work previously appeared in a poster abstract \cite{ACC_VLHCC}.}.
+In this paper we propose a technique called {\it active code completion} that eliminates these restrictions.
 This  makes developing and integrating a broad array of highly-specialized developer tools directly into the editor, via the familiar code completion command, significantly simpler.

### 2014-02-06 thesis-proposal/background.tex
More progress on proposal

@@
 % !TEX root = omar-thesis-proposal.tex
-\subsubsection{Background}\label{alibs}
+\subsubsection{Background}\label{alibs}\label{background}
 The term \emph{active libraries} was first introduced by Veldhuizen et al.
@@
 % This stuff isn't actually hooking into the compiler...
-The first concrete realizations of active libraries, prompting the introduction of the term, were libraries that performed domain-specific program optimization at compile-time by exploiting language mechanisms that allow for limited compile-time computation.
+The first concrete realizations of active libraries in statically typed settings, prompting the introduction of the term, were libraries that performed domain-specific program optimization at compile-time by exploiting language mechanisms that allow for limited compile-time computation.
 A prominent example in the literature is Blitz++, a library that uses C++ template metaprogramming to optimize compound operations on vectors and matrices by eliminating intermediate allocations \cite{veldhuizen2000blitz++}.
@@
 
-More powerful and direct compile-time \emph{term rewriting mechanisms} available in some languages can also be used for optimization, as well as for introducing specialized error checking and reporting\- logic and extending a language with new abstractions.
+More powerful and direct compile-time \emph{term rewriting mechanisms} available in some languages can also be used for optimization, as well as for introducing specialized error checking logic and extending a language with new abstractions.
 These mechanisms are highly expressive because they allow users to programmatically manipulate syntax trees directly, but they suffer from problems of composability and safety.
-For example, compile-time macros, such as those in MetaML \cite{Sheard:1999:UMS}, Template Haskell \cite{SheardPeytonJones:Haskell-02} and Scala \cite{scala-macros}, take full control over all the code that they enclose.
+For example, compile-time macros, such as those in MetaML \cite{Sheard:1999:UMS}, Template Haskell \cite{SheardPeytonJones:Haskell-02} and Scala \cite{scala-macros}, take full control over all of the code that they enclose.
 This can be problematic, however, as outer macros can interfere with the functionality of inner macros.
-Moreover, once a value escapes a macro's scope, there is no way to rely on the guarantees and features that were available within its scope, because the output of a macro is simply a value in the underlying language (a problem fundamentally related to the compatibility problems described in Section \ref{motivation}).
+Moreover, once a value escapes a macro's scope, there is no way to rely on the guarantees and features that were available within its scope, because the output of a macro is simply a value in the underlying language (a problem fundamentally related to the problem of relying on a common intermediate language,  described in Section \ref{external-approaches}).
 Thus, macros can be used to automate code generation, but not to truly extend the semantics of a language.
@@
 
-Some term rewriting systems replace the block scoping of macros with pattern-based dispatch.
+Some term rewriting systems replace the explicitly delimited scoping of macros with global pattern-based dispatch.
 Xroma (pronounced ``Chroma''), for example, is designed around active libraries and allows users to insert custom rewriting passes into the compiler from within libraries \cite{activelibraries}.
-Similarly, the derivative of Haskell implemented by the Glasgow Haskell Compiler (GHC) allows libraries to define custom compile-time term rewriting logic if an appropriate flag is passed in \cite{jones2001playing}.
-In both cases, the user-defined logic can dispatch on arbitrary patterns of code throughout the component or program the extension is activated within, so these mechanisms are highly expressive and avoid some of the difficulties of block-scoped macros.
+Similarly, the derivative of Haskell implemented by the Glasgow Haskell Compiler (GHC) allows providers to introduce custom compile-time term rewriting logic if an appropriate flag is passed in \cite{jones2001playing}.
+In both cases, the user-defined logic can dispatch on arbitrary patterns of code throughout the component or program the extension is activated within, so these mechanisms are highly expressive and avoid some of the difficulties of explicitly invoked macros.
 But libraries containing such global rewriting logic cannot be safely composed because two different libraries may attempt to rewrite the same piece of code differently.
-It is also difficult to guarantee that such logic is correct and difficult to reason about code when importing a library can change the semantics of the program in a non-local manner.
+It is also difficult to guarantee that such logic is correct and difficult to reason about code when simply importing a library can change the semantics of the program in a highly non-local manner.
 % http://www.haskell.org/ghc/docs/latest/html/users_guide/rewrite-rules.html
 
-Another example of an active library approach to extensibility is SugarJ \cite{erdweg2011sugarj} and other generated by Sugar* \cite{erdweg2013framework}, like SugarHaskell \cite{erdweg2012layout}.
-These languages permit libraries to extend the base syntax of the core language in a nearly-arbitrary manner, and these extensions are imported transitively throughout a program.
-These extensions are also not safely composable.
+Another example of an active library approach to extensibility is SugarJ \cite{erdweg2011sugarj} and other languages generated by Sugar* \cite{erdweg2013framework}, like SugarHaskell \cite{erdweg2012layout}.
+These languages permit libraries to extend the base syntax of the core language in a nearly arbitrary manner, and these extensions are imported transitively throughout a program.
+Unfortunately, this flexibility means that  extensions are also not safely composable.
 For example, a library that defines a literal syntax for HTML would conflict with another that defines a literal syntax for XML because they define differing semantics for some of the same syntactic forms.
-If SugarJ was used by two different regular expression engines to provide literal syntax for regular expression patterns, there could easily be conflicts at link-time because both will introduce many of the same standard notations but back them with differing implementations.
-And again, it can be difficult for users to predict what type of data an unfamiliar piece of syntax desugars into, leading to difficulties reading and reasoning about code.
+If SugarJ was used by two different regular expression engines to provide literal syntax for regular expression patterns, there could easily be conflicts at link-time because both will introduce many of the same notations but back them with differing implementations.
+And again, it is difficult to predict what an unfamiliar piece of syntax desugars into, leading to difficulties reading and reasoning about code.
 

### 2014-02-06 thesis-proposal/parsing.tex
More progress on proposal

--- 2013-09-10 thesis-proposal/acc.tex
@@
 % !TEX root = omar-thesis-proposal.tex
-\section{Active Code Completion}
+\lstset{tabsize=2, 
+basicstyle=\ttfamily\fontsize{8pt}{1em}\selectfont, 
+commentstyle=\itshape\rmfamily, 
+stringstyle=\ttfamily,
+numbers=left, numberstyle=\scriptsize\color{gray}\ttfamily, language=ML,moredelim=[il][\sffamily]{?},mathescape=true,showspaces=false,showstringspaces=false,xleftmargin=15pt,escapechar=@, morekeywords=[1]{tyfam,opfam,let,fn,val,def,casetype,objtype,metadata,of,*,datatype,new,valAST},deletekeywords={for,double,in},classoffset=0,belowskip=\smallskipamount,
+moredelim=**[is][\color{cyan}]{SSTR}{ESTR},
+moredelim=**[is][\color{OliveGreen}]{SHTML}{EHTML},
+moredelim=**[is][\color{purple}]{SCSS}{ECSS},
+moredelim=**[is][\color{brown}]{SSQL}{ESQL},
+moredelim=**[is][\color{orange}]{SCOLOR}{ECOLOR},
+moredelim=**[is][\color{magenta}]{SPCT}{EPCT}, 
+moredelim=**[is][\color{gray}]{SNAT}{ENDNAT}, 
+moredelim=**[is][\color{blue}]{SURL}{EURL},
+moredelim=**[is][\color{SeaGreen}]{SQT}{EQT},
+moredelim=**[is][\color{Periwinkle}]{SGRM}{EGRM},
+moredelim=**[is][\color{YellowGreen}]{SID}{EID}
+}
+\lstloadlanguages{Java,VBScript,XML,HTML}
+\let\li\lstinline
+
+\section{Type-Specific Languages}\label{aparsing}
+General-purpose abstraction mechanisms are quite powerful.
+By using a general-purpose abstraction mechanism to encode a data structure, one immediately inherits a body of primitive operations, established reasoning principles, well-optimized implementations and tool support.
+For example, lists can be encoded using inductive datatypes, the general-purpose abstraction mechanism that most functional programming languages emphasize.
+Intuitively, a list can either be empty, or be broken down into a \emph{head} element and a \emph{tail}, another list.
+In an ML-like language, the polymorphic list type is declared:
+\begin{lstlisting}[numbers=none]
+datatype 'a list = Nil | Cons of 'a * 'a list
+\end{lstlisting}
+By encoding lists in this way, we can immediately reason about them by structural induction, introduce them by naming the appropriate case and providing the associated data, if any, and examine them by pattern matching.
+%The programmer only needs to provide an encoding of the structure of lists; the semantics and implementation are filled in by the general-purpose abstraction mechanism.
+
+While inheriting these operations and reasoning principles can be quite useful, inheriting the associated general-purpose syntax can sometimes be a liability.
+For example, few would claim that writing a list of numbers as a sequence of \li{Cons} cells is convenient:
+\begin{lstlisting}[numbers=none]
+Cons(1, Cons(2, Cons(3, Cons(4, Nil))))
+\end{lstlisting}
+
+Because lists are a common data structure, many languages include specialized syntax for introducing them, e.g.
+\li{[1, 2, 3, 4]}.
+This notation is semantically equivalent to the general-purpose notation shown above, but brings cognitive benefits by drawing attention to the content of the list, rather than the nature of the encoding.
+To use terminology from the literature on the cognitive dimensions of notations \cite{green1996usability}, it is more \emph{terse}, \emph{visible} and \emph{maps more closely} to the intuitive notion of a list.
+
+Although list, number and string literals are nearly ubiquitous features of modern languages, some languages also  provide specialized notation for other common data structures (like maps and sets), data formats (like XML and JSON), query languages (like regular expressions and SQL), markup languages (like HTML) and others.
+For example, a language with built-in notation for HTML and SQL, with type-safe interpolation of host language terms within curly braces, might allow:
+\begin{lstlisting}
+let webpage : HTML = SHTML<html><body><h1>Results for {EHTMLkeywordSHTML}</h1>
+  <ul id="results">{EHTMLto_list_items(query(db, 
+    SSQLSELECT title, snippet FROM products WHERE {ESQLkeywordSSQL} in titleESQL)SHTML}
+  </ul></body></html>EHTML
+\end{lstlisting}
+to be shorthand for:
+\begin{lstlisting}
+let webpage : HTML = HTMLElement(Dict.empty(), [BodyElement(Dict.empty(),
+  [H1Element(Dict.empty(), [TextNode(concat($\texttt{"}$SSTRResults for $\texttt{"}$ESTR, keyword))]), 
+  ULElement((Dict.add Dict.empty() ($\texttt{"}$SSTRid$\texttt{"}$ESTR, $\texttt{"}$SSTRresults$\texttt{"}$ESTR)), to_list_items(query(db, 
+    SelectStmt([$\texttt{"}$SSTRtitle$\texttt{"}$ESTR, $\texttt{"}$SSTRsnippet$\texttt{"}$ESTR], $\texttt{"}$SSTRproducts$\texttt{"}$ESTR, 
+      [WhereClause(InPredicate(StringLit(keyword), $\texttt{"}$SSTRtitle$\texttt{"}$ESTR))]))))]]]
+\end{lstlisting}
+
+When a specialized notation like this is not available, but the equivalent general-purpose notation is too cognitively demanding for comfort, developers typically turn to run-time mechanisms to make constructing data structures more convenient.
+Among the most common strategies in these situations is to simply use a string representation that is parsed at run-time.
+Developers across language paradigms frequently write examples like those above as:
+\begin{lstlisting}
+let webpage : HTML = parse_html($\texttt{"}$SSTR<html><body><h1>Results for ESTR$\texttt{"}$ + keyword + $\texttt{"}$SSTR</h1>
+  <ul id=\$\texttt{\color{cyan}"}$results\$\texttt{\color{cyan}"}$>$\texttt{"}$ESTR + to_string(to_list_items(query(db, parse_sql(
+  	$\texttt{"}$SSTRSELECT title, snippet FROM products WHERE '$\texttt{" + keyword + "}$' in title$\texttt{"}$ESTR)))) + 
+  $\texttt{"}$SSTR</ul></body></html>$\texttt{")}$
+\end{lstlisting}
+
+Though recovering much of the notational convenience of the literal version, it is still more awkward to write, requiring explicit conversions to and from structured representations (\li{parse_html} and \li{to_string}, respectively) and escaping when the syntax of the language interferes with the syntax of string literals (line 2).
+Code like this also causes a number of problems beyond cognitive load.
+Because parsing occurs at run-time, syntax errors will not be discovered statically, causing potential problems in production scenarios.
+Run-time parsing also incurs performance overhead, particularly relevant when code like this is executed often (as on a heavily-trafficked website).
+But the most serious issue with this code is that it is fundamentally insecure: it is vulnerable to cross-site scripting attacks (line 1) and SQL injection attacks (line 3).
+For example, if a user entered the keyword \li{'; DROP TABLE products --}, the entire product database could be erased.
+These attack vectors are considered to be two of the most serious security threats on the web today \cite{owasp2013}.
+Although developers are cautioned to sanitize their input, it can be difficult to verify that this was done correctly throughout a codebase.
+The most straightforward way to avoid these problems today is to insist on structured representations, despite their inconvenience.
+
+Unfortunately, it does not appear that this is sufficient.
+Situations like this, where maintaining strong correctness, performance and security guarantees entails significant syntactic overhead, causing developers to turn to worse solutions that are more convenient, are quite common.
+To emphasize this, let us return to our running example of pattern literals.
+A small regular expression like \verb!(\d\d):(\d\d)\w*((am)|(pm))!
+might be written using general-purpose notation as:
+\begin{lstlisting}
+Seq(Group(Seq(Digit, Digit), Seq(Char($\texttt{"}$SSTR:ESTR$\texttt{"}$), Seq(Group(Seq(Digit, Digit)), 
+  Seq(ZeroOrMore(Whitespace), Group(Or(Group(Seq(Char($\texttt{"}$SSTRaESTR$\texttt{"}$), Char($\texttt{"}$SSTRmESTR$\texttt{"}$))), 
+  Group(Seq(Char($\texttt{"}$SSTRpESTR$\texttt{"}$), Char($\texttt{"}$SSTRmESTR$\texttt{"}$))))))))))
+\end{lstlisting}
+This is clearly more cognitively demanding, both when authoring the regular expression and when reading it.
+Among the most common strategies in these situations, for users of both object-oriented and functional languages, is to simply use a string representation that is parsed at run-time.
+\begin{lstlisting}
+rx_from_str($\texttt{"}$SSTR(\\d\\d):(\\d\\d)\\w*((am)|(pm))ESTR$\texttt{"}$)
+\end{lstlisting}
+%%
+%%For example, in languages without SQL literals, developers can implement a builder pattern:
+%%\begin{lstlisting}
+%%new SQLQuery().SELECT("*").FROM("table").WHERE("username").Eq(username)
+%%\end{lstlisting}
+This is problematic, for all of the same reasons as described above: needing explicit conversions between representations, interference issues with string syntax, correctness problems, performance overhead and security issues.
+
+Today, supporting new literal notations within an existing language requires the cooperation of the language designer.
+This is primarily because, with conventional parsing strategies, not all notations can unambiguously coexist, so a designer is needed to make choices about which syntactic forms are available and what their semantics should be.
+For example, conventional notations for sets and maps are both delimited by curly braces.
+When Python introduced set literals, it chose to distinguish them based on whether the literal contained only values (e.g.
+\verb|{3}|), or key-value pairs (\verb|{"x": 3}|).
+But this causes an ambiguity with the syntactic form \verb|{ }| -- should it mean an empty set or an empty map (called a dictionary in Python)?
+The designers of Python chose the latter interpretation (for backwards compatibility reasons).
+
+So although languages that allow users to introduce new syntax from within libraries appear to hold promise for the reasons described above, providing this form of extensibility is non-trivial because there is no longer a central designer making decisions about such ambiguities.
+In most existing related work, the burden of resolving ambiguities falls to the clients of extensions.
+For example, SugarJ \cite{erdweg2011sugarj} and other extensible languages generated by Sugar* \cite{erdweg2013framework} allow providers to extend the base syntax of the host language with new forms, like set and map literals.
+These new forms are imported transitively throughout a program.
+To resolve syntactic ambiguities that arise, clients must manually augment the composed grammar with new rules that allow them to choose the correct interpretation explicitly.
+This is both difficult to do, requiring a reasonably thorough understanding of the underlying parser technology (in Sugar*, generalized LR parsing) and increases the cognitive load of using the conflicting notations (e.g.
+both sets and dictionaries) in the same file.
+These kinds of conflicts occur in a variety of circumstances: HTML and XML, different variants of SQL, JSON literals and dictionaries, or simply different implementations (``desugarings'') of the same specialized syntax (e.g.
+two regular expression engines) can all cause problems.
+
+In this work, we will describe an alternative parsing strategy that avoids these problems by shifting responsibility for parsing certain \emph{generic literal forms} into the typechecker.
+The typechecker, in turn, defers responsibility to user-defined types, by treating the body of the literal as a term of the   \emph{type-specific language (TSL)} associated with the type it is being checked against.
+The TSL is responsible for rewriting this term to ultimately use only general-purpose notation.
+This strategy avoids the problem of conflicting syntax, because neither the base language nor TSLs are ever extended directly.
+It also avoids semantic conflicts -- the meaning of a form like \verb|{ }| can differ depending  on its type, so it is safe to use it for empty sets, dictionaries and other data structures, like JSON literals.
+This also frees these common notations from being tied to the variant of a  data structure built into the standard library, which sometimes does not provide the exact semantics that a programmer needs (for example, Python dictionaries do not preserve key insertion order).
+%\begin{lstlisting}
+%let empty_set : Set = { }
+%let empty_dict : Dict = { }
+%let empty_json : JSON = { }
+%\end{lstlisting}
+\subsection{Wyvern}
+We develop our work as a variant of a new programming language being developed by our group called Wyvern \cite{Nistor:2013:WST:2489828.2489830}.
+To allow us to focus on the essence of our proposal, the variant of Wyvern we will describe in this thesis is simpler than the variant previously described: it is purely functional (there are no effects or mutable state) and it does not enforce a uniform access principle for objects (fields can be accessed directly).
+Objects are thus essentially labeled product types with simple methods (functions that require a self-reference).
+We also add recursive labeled sum types, which we call \emph{case types}, that are quite similar to datatypes in ML.
+One can refer to the version of the language described in this thesis as \emph{TSL Wyvern} when the variant is not clear.
+
+\subsection{Example: Web Search}
+We begin in Fig.
+\ref{f-example} with an example showing several different TSLs being used to define a fragment of a web application showing search results from a database.
+We will go through this example below to develop intuitions about TSLs in Wyvern.
+Note that for clarity of presentation, we color each character  according to the TSL it is governed by.
+\begin{figure}[t]
+\begin{lstlisting}
+let imageBase : URL = <SURLimages.example.comEURL>
+let bgImage : URL = <SURL%EURLimageBaseSURL%/background.pngEURL>
+new : SearchServer
+  def resultsFor(searchQuery : String, page : Nat) : Unit =
+    serve(~) (* serve : HTML -> Unit *)
+SHTML      :html
+        :head
+          :title Search Results
+          :style EHTML~
+SCSS            body { background-image: url(%ECSSbgImageSCSS%) }
+            #search { background-color: %ECSS`SCOLOR#aabbccECOLOR`.darken(SPCT20pctEPCT)SCSS% }
+ECSSSHTML        :body
+          :h1 Results for {EHTMLsearchQuerySHTML}
+          :div[id="search"]
+            Search again: {EHTMLSearchBox($\texttt{"}$SSTRGo!ESTR$\texttt{"}$)SHTML}
+          {EHTML (* fmt_results : DB * SQLQuery * Nat * Nat -> HTML *)
+            fmt_results(db, ~, SNAT10ENDNAT, page)
+              SSQLSELECT * FROM products WHERE {ESQLsearchQuerySSQL} in titleESQL
+          SHTML}
+\end{lstlisting}
+\vspace{-8px}
+\caption{Wyvern Example with Multiple TSLs}
+\label{f-example}
+\vspace{-10px}
+\end{figure}
+\subsection{Inline Literals}
+\begin{figure}[t]
+\begin{lstlisting}
+<SURLliteral body here, <inner angle brackets> must be balancedEURL>
+{SURLliteral body here, {inner braces} must be balancedEURL}
+[SURLliteral body here, [inner brackets] must be balancedEURL]
+`SURLliteral body here, ``inner backticks`` must be doubledEURL`
+$\texttt{'}$SURLliteral body here, ''inner single quotes'' must be doubledEURL$\texttt{'}$
+$\texttt{"}$SURLliteral body here, ""inner double quotes"" must be doubledEURL$\texttt{"}$
+SURL12xyzEURL (* no delimiters necessary for number literals; suffix optional *)
+\end{lstlisting}
+\vspace{-8px}
+\caption{Inline Generic Literal Forms}
+\vspace{-10px}
+\label{f-delims}
+\end{figure}
+Our first TSL appears on the right-hand side of the variable binding on line 1.
+The variable \li{imageBase} is annotated with its type, \li{URL} (not shown).
+This is an {object type} declaring several fields representing the components of a URL: its protocol, domain name, port, path and so on.
+We could have created a value of type \li{URL} using general-purpose notation\todo{fix comment formatting}\todo{new color}:
+\begin{lstlisting}
+let imageBase : URL = new
+  val protocol : URLString = $\text{"}$SSTRhttpESTR$\texttt{"}$
+  val subdomain : URLString = $\texttt{"}$SSTRimagesESTR$\texttt{"}$
+  (* ... *)
+\end{lstlisting}
+%  val domain : URLString = $\texttt{"}$SSTRexampleESTR$\texttt{"}$
+This is tedious.
+Instead, because the \li{URL} type has a TSL associated with it, we can instead introduce precisely this value using conventional notation for URLs by placing it in the \emph{body} of a \emph{generic literal}, \li{<SURLimages.example.comEURL>}.
+Any other delimited form in Fig.
+\ref{f-delims} could equivalently be used if the constraints shown are obeyed.
+The type annotation on \li{imageBase} implies that this literal's \emph{expected type} is \li{URL}, so the {body} of the literal (the characters between the angle brackets, in blue) will be governed by the \li{URL} TSL during the typechecking phase.
+This TSL will parse the body ({at compile-time}) to produce a Wyvern abstract syntax tree (AST) that explicitly instantiates a new object of type \li{URL}.
+
+In addition to supporting conventional notation for URLs, this TSL supports \emph{typed interpolation}
+of a Wyvern expression of type \li{URL} to form a larger URL. The interpolated term is delimited by percent signs,
+as seen on line 2 of Fig.
+\ref{f-example}.
+The TSL parses code between percent signs  as a Wyvern expression with expected type \li{URL}, using its AST to construct an AST for the expression as a whole.
+String interpolation
+never occurs.
+Note that the delimiters that are used to go from Wyvern to a TSL are controlled by Wyvern (those in Fig.
+\ref{f-delims}) while the delimiters that can be used to go from a TSL back to Wyvern are controlled by the TSL.
+
+%\subsection{General-Purpose Notation for Object and Case Types}
+%The general-purpose introductory form for object types is \li{new}.
+This form is a syntactic \emph{forward reference} to the layout-delimited block of {definitions} beginning on the line immediately after the line \li{new} appears in (line 4 in this case), and ending when the indentation level has returned to the baseline, or when the file ends (after line 19 in this case).
+An object in TSL Wyvern can contain methods, introduced using \li{def}, and fields, introduced using \li{val}.
+Here we have just a single method, \li{serve_results} taking two arguments.
+Object types in TSL Wyvern are simple structural interfaces that constrain the signatures of fields and methods.
+The \emph{type ascription} around \li{new} checks that the object being introduced satisfies the signature of \li{SearchServer} (not shown).\todo{discuss case types}
+\subsection{Layout-Delimited Literals}
+On line 5 of Fig.
+\ref{f-example}, we see a call to a function \li{serve}, not shown, which has type \li{HTML -> Unit}.
+Here, \li{HTML} is a user-defined \emph{case type}, having cases for each HTML tag as well as some other structures, like text nodes and sequencing.
+Declarations  of some of these cases can be seen on lines 2-6 of Fig.
+\ref{f-htmltype} (note that TSL Wyvern also includes simple product types for convenience, written \li{T1 * T2}).
+We could again use Wyvern's general-purpose introductory form for case types, e.g.
+\li{HTML.BodyElement((attrs, child))} ({unlike in ML, in Wyvern we must explicitly qualify constructors with the case type they are part of when they are used}.
+This is largely to make our formal semantics simpler and for clarity of presentation.) But, as discussed above, using this syntax can be inconvenient and  cognitively demanding.
+Thus, we associate a TSL with \li{HTML} that provides a simplified notation for writing HTML, shown being used on lines 6-20 of Fig.
+\ref{f-example}.
+This literal body is layout-delimited, rather than delimited by explicit tokens as in Fig.
+\ref{f-delims}, and introduced by a form of \emph{forward reference}, written \li{~} (``tilde''), on the previous line.
+Because the forward reference occurs in a position -- as the argument to the function \li{serve} -- where the expected type is \li{HTML}, the literal body is governed by that type's TSL.
+The forward reference will be replaced by the general-purpose term, of type \li{HTML}, generated by the TSL during typechecking.
+\subsection{Implementing a TSL}
+Portions of the implementation of the TSL for \li{HTML} are shown on lines 8-17 of Fig.
+\ref{f-htmltype}.
+A TSL is associated with a type, forming an \emph{active type}, using a more general mechanism for associating a value with a type.
+A value associated with a type is called the type's  metadata, and is introduced as shown on line 8 of Fig.
+\ref{f-htmltype}.
+For the purposes of this work, metadata values will always be of type \li{HasTSL}, an object type that declares a single field, \li{parser}, of type \li{Parser}.
+The \li{Parser} type is an object type declaring a single method, \li{parse}, that transforms a \li{TokenStream} to a Wyvern AST, which is a value of type \li{Exp}.
+This is in turn a case type that encodes the syntax of Wyvern expressions.
+Fig.
+\ref{f-builtins} shows these types.
+
+Notice, however, that the TSL for \li{HTML} is not provided as an explicit \li{parse} method but instead as a declarative grammar.
+A grammar is a specialized notation for defining a parser, so we can implement a more convenient grammar-based parser generator  as a TSL associated with the \li{Parser} type.
+We chose the  layout-sensitive formalism developed by Adams \cite{Adams:2013:PPI:2429069.2429129} -- Wyvern is itself layout-sensitive and has a grammar that can be written down using this formalism, so it is sensible to expose it to TSL providers as well.
+Most aspects of this formalism are completely conventional.
+Each non-terminal (e.g.
+\li{start}) is defined by a number of disjunctive productions, each introduced using \li{->}.
+Each production defines a sequence of terminals (e.g.
+\li{':body'}) and non-terminals (e.g.
+\li{start}, or one of the built-in non-terminals \li{ID}, \li{EXP} or \li{TYPE}, representing Wyvern identifiers, expressions and types, respectively).
+Unique to Adams' formalism is that each terminal and non-terminal in a production can also have an optional \emph{layout constraint} associated with it.
+The layout constraints available are \li{=} (meaning that the leftmost column of the annotated term must be aligned with that of the parent term), \li{>} (the leftmost column must be indented further) and \li{>=} (the leftmost column \emph{may} be indented further).
+We will discuss this formalism further when we formally specify Wyvern's layout-sensitive concrete syntax.
+
+Each production is followed, in an indented block, by a Wyvern function that generates a value given $n$ arguments, which will bound to the values recursively generated by each of the $n$ non-terminals it contains, ordered left-to-right.
+For the starting non-terminal, always called \li{start}, this function should return a value of type \li{Exp}.
+Here, we show how to generate an AST using general-purpose notation for \li{Exp} (lines 13-15) as well as a more natural \emph{quasiquote} style (lines 11 and 18).
+Quasiquotes are expressions that are not evaluated, but rather reified into their corresponding syntax tree.
+We observe that quasiquotes too fall into the pattern of ``specialized notation associated with a type'' -- quasiquotes for expressions, types and identifiers are simply TSLs associated with \li{Exp}, \li{Type} and \li{ID} (Fig.
+\ref{f-builtins}).
+They support the full Wyvern concrete syntax as well as an additional delimited form, written with \li{%}s, that supports ``unquoting'': interpolating another AST into the one being generated.
+Again, interpolation is type safe and structured, rather than based on string interpolation.
+This is accomplished by placing the interpolated value in a place in the generated AST where its type is known -- on line 11 of Fig.
+\ref{f-htmltype} it is known to be \li{HTML} and on line 13 it is known to be \li{CSS} by the declaration of \li{HTML}, and on line 15, it is known to be \li{HTML} by the use of an explicit ascription.
+When these generated ASTs are recursively typechecked during compilation, any use of a nested TSL at the top-level (e.g.
+the CSS TSL in Fig \ref{f-example}) will operate as intended.
+
+\begin{figure}
+\begin{lstlisting}[escapechar=$]
+casetype HTML 
+    Empty of Unit
+  | Text of String
+  | Seq of HTML * HTML 
+  | BodyElement of Attributes * HTML
+  | StyleElement  of Attributes * CSS
+  | ...
+  metadata = new : HasTSL
+    val parser : Parser = ~
+SGRM      start -> ':body'= start>
+        EGRMfn child:Exp => `SQTHTML.BodyElement(([], %EQTchildSQT%))EQT`
+SGRM      start -> ':style'= EXP>
+        EGRMfn e:Exp => `SQTHTML.StyleElement(([], %EQTeSQT%))EQT`
+SGRM      start -> '{'= EXP '}'=EGRM
+        fn e:Exp => `SQT%EQTeSQT% : HTMLEQT`
+\end{lstlisting}
+\vspace{-8px}
+\caption{A Wyvern case type with an associated TSL.}
+\label{f-htmltype}
+\end{figure}
+\begin{figure}[t]
+\begin{subfigure}[t]{.53\textwidth}
+\begin{lstlisting}
+objtype HasTSL
+  val parser : Parser
+objtype Parser                          
+  def parse(ts : TokenStream) : (Exp * 
+    TokenStream)
+  metadata = new : HasTSL
+    val parser : Parser = (* parser generator *)      
+casetype Type
+    Var of ID
+  | Arrow of Type * Type
+  | Prod of Type * Type
+  metadata = new : HasTSL
+    val parser : Parser = (* type quasiquotes *)
+\end{lstlisting}
+\end{subfigure}
+\begin{subfigure}[t]{.55\textwidth}
+\begin{lstlisting}[linewidth=.5\textwidth]
+casetype Exp 
+    Var of ID
+  | Lam of ID * Type * Exp
+  | Ap of Exp * Exp
+  | ProdIntro of Exp * Exp
+  | CaseIntro of Type * ID * Exp
+  ... 
+  | FromTS of Exp * Exp
+  | Literal of TokenStream
+  | Error of ErrorMessage
+  metadata = new : HasTSL
+    val parser : Parser = (* quasiquotes *)
+\end{lstlisting}
+\end{subfigure}
+\caption{Some of the types included in the Wyvern prelude.}
+\vspace{-10px}
+\label{f-builtins}
+%\label{fig:typeParser}
+\end{figure}
+
+\subsection{Formalization}
+A formal and more detailed description can be found in our paper draft\footnote{\url{https://github.com/wyvernlang/docs/tree/master/ecoop14}}. In particular:
+\begin{enumerate}
+\item We provide a more complete layout-sensitive concrete syntax.
+We show how it can be written without the need for a context-sensitive lexer or parser and give a full specification for the layout-delimited literal form introduced by a forward reference, \li{~}, as well as other forms of forward-referenced blocks.
+%\item We detail the general mechanism for associating metadata with a type.
+A TSL is then implemented by associating a parser (of type \li{Parser}) with a type.
+The parser is responsible for rewriting tokenstreams (of type \li{Tokenstream}) into Wyvern ASTs (of type \li{Exp}).
+These types are defined in the standard library.
+%\item This lower-level mechanism is general, but writing a hand-written parser and manipulating syntax trees manually is cognitively demanding.
+We observe that \emph{grammars} and \emph{quasiquotes} can both be seen as TSLs for parsers and ASTs respectively and discuss how to implement them as such.
+\item We formalize the static semantics, including the literal parsing logic, of TSL Wyvern by combining a bidirectional type system (in the style of Lovas and Pfenning \cite{lovasPfenning}) with an elaboration semantics (in the style of Harper and Stone \cite{harperStone}).
+By distinguishing locations where an expression synthesizes a type from locations where an expression is being analyzed against a known type, we can precisely state where generic literals can and cannot appear and how parsing is delegated to a TSL.
+\item A na\"ive rewriting strategy would be \emph{unhygienic} -- it could allow for the inadvertent capture of local variables.
+We show a novel mechanism that ensures hygiene by requiring that the generated AST is closed except for subtrees derived from portions of the user's tokenstream that are interpreted as nested Wyvern expressions.
+We also show how to explicitly refer to local values available in the parser definition (e.g.
+helper functions) in a safe way.
+We formalize this hygiene mechanism by bifurcating the context in our static semantics.
+\item We provide several examples of TSLs throughout the paper, but to examine how broadly applicable the technique is, we conduct a simple corpus analysis, finding that string languages are used ubiquitously in existing Java code  (collaborative work with Darya Kurilova\todo{and Josh?}).
+\end{enumerate}
+
+\subsection{Remaining Tasks and Timeline}
+The following tasks remain to be completed. I have scheduled this for the period of March 2nd-25th, before the OOPSLA submission deadline.
+\begin{enumerate}
+\item In our current top-level grammar, expressions within parenthesis must still obey layout constraints.
+This is not strictly necessary for preventing ambiguities, and it is quite convenient to not require this (as in Python, for example).
+We will add these simple rules to the grammar.
+\item We must write down the full formal semantics (including rules that we omitted for concision in the paper draft) in a technical report, as well as provide more complete proofs of the metatheory. 
+\item Our current static semantics does not support mutually recursive type declarations, but this is necessary for our prelude to typecheck, so we will add support for this.
+\item We must further consider aspects of hygiene.
+In particular, we do not yet have a clean mechanism for preventing unintentional variable \emph{shadowing}, only unintentional variable \emph{capture}.
+It may be possible to prevent shadowing by making it impossible for variables to be introduced into interpolated Wyvern expressions (so that function application is the only way to pass data from a TSL to Wyvern code, as in the \li{Parser} TSL we have shown).
+\item The corpus analysis we conducted was preliminary. We must perform this in a more complete and rigorous manner (collaborative work with Darya Kurilova).
+%\item The current implementation of Wyvern does not include all of these mechanisms as described.
+Although the implementation of Wyvern is not a project I am responsible for, I plan on working with the student who is doing this to implement as much of what we have described as possible.\todo{should I drop this from the proposal?
+I don't want to promise that everything will be implemented because that is largely contingent on Benjamin's effort, not mine.}
+\end{enumerate}
+%
+%\subsection{TODO}
+%- Figure out separate compilation (e.g. of CSS)

### 2014-02-06 thesis-proposal/problem.tex
More progress on proposal

--- 2014-01-11 thesis-proposal/problem.tex
@@
 \section{Motivation}\label{motivation}
-Specifying and implementing a programming language together with its supporting tools (collectively, a \emph{programming system}) that has sound theoretical foundations, helps users identify and fix errors as early as possible, supports a natural programming style, and performs well across diverse problem domains and hardware platforms remains a grand challenge in computing.
-Due to the increasing diversity and complexity of modern problem domains and hardware platforms, it has become clear that no small  collection of general-purpose constructs backed by a simple, all-purpose implementation can fully satisfy these criteria in all cases.
-Instead,  researchers and domain experts (collectively, \emph{providers}) continue to develop  specialized notations,  static semantics, dynamic semantics, implementation strategies, optimizations, run-time systems and tools (collectively, \emph{features})  designed to address these modern challenges in different circumstances.
-%, researchers and domain experts must continue to design and develop, in a principled manner, specialized notations, type systems, optimizations, run-time systems and tools (collectively, \emph{features}) to support developers as they work in increasingly complex problem domains and on modern hardware systems.
-
-% Rather, researchers and domain experts must continue, in a principled manner, to specify and implement novel abstractions, provide alternative implementations of existing abstractions, supplement languages with useful tools and improve tools with new behaviors (collectively, realize new \emph{features}) that better satisfy these criteria by balancing the concerns relevant to particular application domains.
-
-Ideally, a provider would develop and distribute such new features orthogonally, as libraries, so that client developers could granularly choose those that best satisfy their needs.
-Unfortunately this is often infeasible because libraries are vehicles for specifying a program in terms of a monolithic collection of existing features.
-From the perspective of a library, the language's syntax, type system and semantics are fixed in advance, the compiler and run-time system are ``black box'' implementations of this fixed language, and the other tools, like code editors and debuggers, operate according to domain-agnostic protocols also based only on a fixed language specification.
-As a result, providers of new system features must often take \emph{language-external approaches}, sometimes designing a new system altogether.
-We will argue that such approaches are problematic, and that taking them has led to an unnecessary gap between research and practice.
-In place of these approaches, we will advocate for \emph{language-integrated extensibility mechanisms}, and show how, by organizing new features around {types} enforcing certain constraints, these mechanisms can be made both safe and highly expressive.
-We call a type that introduces new features an \emph{active type} and programming systems organized around active types \emph{actively-typed programming systems}.
+Specifying and implementing a programming language together with its supporting tools (collectively, a \emph{programming system}) that is built upon sound theoretical foundations, helps users identify and fix errors as early as possible, supports a natural programming style, and performs well across diverse problem domains and hardware platforms remains a grand challenge in computing.
+In view of this goal, researchers and domain experts (collectively, \emph{providers}) continue to develop new special-purpose syntax, static and dynamic semantics, implementation strategies, optimizations, run-time systems and tools (collectively, \emph{features})  designed to address these challenges in increasingly diverse contexts.
+Ideally, a provider would be able to develop and distribute these kinds of new features orthogonally, as libraries, so that client developers could granularly choose those that best satisfy their needs.
+Unfortunately this is often infeasible because from the perspective of a library, the language's syntax and semantics are fully specified in advance, the compiler and run-time system are ``black box'' implementations of this fixed specification, and the other tools, like code editors and debuggers, operate according to fixed protocols.
+Providers of new system features must, as a result, take \emph{language-external approaches}, often by deriving a new programming system altogether.
+This approach has been encouraged, historically, by the availability of tools like compiler generators and language workbenches.
+We will argue that these approaches are technically problematic and that taking them has led to an unnecessary gap between research and practice.
+In their place, we will develop \emph{language-integrated extensibility mechanisms} that decentralize control over several core aspects of the programming system.
+By organizing new features around {types} and constraining them appropriately, we aim to show that these mechanisms can guarantee safety and non-interference of extensions while also being highly expressive.
 %But taking a \emph{language-internal approach} to implementing a feature is the most practical.
@@
 \subsection{Motivating Example: Regular Expressions}\label{regex}
-To make the issue concrete, let us begin with a simple example.
-\emph{Regular expressions} are a widely-used mechanism for finding patterns in semi-structured strings (e.g.
+To make the problem we aim to address concrete, we begin with a simple example that we will return to throughout this work.
+\emph{Regular expressions} are a widely-used abstraction for finding patterns in semi-structured strings (e.g.
 DNA sequences) \cite{Thompson:1968:PTR:363347.363387}.
-A programming system that included full compile-time and edit-time support for regular expressions might simultaneously provide features like these:
+If a programming system wished to fully support regular expressions, it might simultaneously provide features like these:
 
@@
 \item \textbf{Built-in syntax for pattern literals} (e.g.
-\cite{perlre}) so that malformed patterns result in intelligible \emph{compile-time} parsing errors.
-In languages lacking such literals, run-time errors relating to pattern syntax can occur, even in well-tested, deployed code \cite{spishak2012type}.
-\item \textbf{Typechecking logic} that ensures that key constraints related to regular expressions are not violated at compile-time:
+\cite{perlre}) so that the cognitive load of using regular expressions is low and malformed patterns result in intelligible {compile-time} errors.
+\item A \textbf{static and dynamic semantics} that ensures, at compile-time whenever possible, that key invariants related to regular expressions are maintained:
 	\begin{enumerate}
@@
 	\item out-of-bounds backreferences are not used \cite{spishak2012type}
-	\item strings known to be in the language of a regular expression are given a type that tracks this information and ensures that string manipulation operations do not inadvertently lead to a string that is not in the assumed language \cite{fulton-thesis}.
+	\item strings known to be in the language of a regular expression are given a type that tracks this information to ensure that string manipulation operations do not inadvertently lead to a string that is not in an assumed language \cite{fulton-thesis}.
 	\end{enumerate}
 When a type error is found, an intelligible error message is provided.
-\item \textbf{Elaboration logic} that partially or fully compiles known regular expressions into the efficient internal representation that will be used by the regular expression matching engine (e.g., a finite automata \cite{Thompson:1968:PTR:363347.363387}) ahead of time.
+%\item An \textbf{implementation} that partially or fully compiles known regular expressions into the efficient internal representation that will be used by the regular expression matching engine (e.g., a finite automata \cite{Thompson:1968:PTR:363347.363387}) ahead of time.
 In most languages, this compilation step occurs at run-time, even if the pattern is fully known at compile-time, thereby introducing performance overhead into programs.
@@
 %By performing this step ahead-of-time, these dangers can be avoided.
-\item \textbf{Editor services} for quickly testing regular expression patterns against test strings,  referring to documentation or searching databases of common patterns (e.g. \cite{IntelliJRegexp}).
+\item \textbf{Editor services} that allow clients to interactively test regular expression patterns against test strings, refer to documentation or search for common patterns (e.g.
+\cite{IntelliJRegexp}).
 \end{enumerate}
@@
 %In a conventional \emph{monolithic} programming system, support for each of these features would need to be built into the language and tools. 
-No system today has built-in support for all of the features enumerated above.
-Instead, libraries generally provide support for regular expressions by leveraging  general-purpose constructs.
-Unfortunately, it is impossible to fully encode the syntax and the specialized static and dynamic semantics described above in terms of general-purpose notations and abstractions like objects or inductive datatypes.
-Library providers have thus needed to compromise, typically by asking clients to provide regular expressions as strings, thereby deferring parsing, typechecking and elaboration to run-time, which introduces performance overhead and can lead to unanticipated run-time errors (as shown in \cite{spishak2012type}) and security vulnerabilities (due to injection attacks when user inputs are spliced into patterns, for example).
-Similarly, useful tools for working with regular expressions are rarely integrated into editors, and even more rarely in a way that facilitates their discovery and use directly when the developer is manipulating regular expressions.
-Tools that must be discovered independently and accessed externally are used less frequently  and can be more awkward than necessary \cite{Campbell:2008:DRT:1636642.1636651, Omar:2012:ACC:2337223.2337324}, leading to lower productivity \cite{Omar:2012:ACC:2337223.2337324}.
+No system today builds in support for all of the features enumerated above.
+Instead, libraries generally provide support for regular expressions by leveraging  general-purpose abstraction mechanisms.
+Unfortunately, it is impossible to fully encode the syntax and the specialized static and dynamic semantics described above in terms of general-purpose notations and abstractions.
+Library providers thus need to compromise, typically by asking clients to provide regular expressions as strings and deferring parsing, typechecking and compilation to run-time.
+This introduces performance overhead and can lead to unanticipated run-time errors (as shown in \cite{spishak2012type}) and security vulnerabilities (due to injection attacks when user inputs are spliced into patterns, for example) \cite{string-stuff}.
+Useful tools for working with regular expressions are rarely integrated into editors, and even more rarely in a way that facilitates their discovery and use directly when the developer is manipulating regular expressions.
+Tools that must be discovered independently and accessed externally are used infrequently \cite{emersonStudy} and can be more awkward than necessary \cite{Campbell:2008:DRT:1636642.1636651, Omar:2012:ACC:2337223.2337324}, leading to lower productivity \cite{emersonStudy, Omar:2012:ACC:2337223.2337324}.\todo{complete references}
 
-\subsection{Language-External Approaches}
+\subsection{Language-External Approaches}\label{external-approaches}
 \begin{figure}
@@
 \label{approaches}
+%\vspace{-10px}
 \end{figure}
 
-When the semantics or implementation of a system must be extended to fully realize a new feature, as in the example above, providers typically take a \emph{language-external approach}, either by developing a new or derivative programming system (supported by \emph{language workbenches} \cite{erdweg2013state},  \emph{DSL frameworks} \cite{fowler2010domain} or \emph{compiler generators}), or by extending an existing system by some mechanism that is not part of the language itself, such as an extension mechanism for a {particular} compiler\footnote{Compilers that modify, or allow modification of, the semantics of their base language, rather than simply permitting meaning-preserving optimizations, should be considered a pernicious means for creating new languages.
-Many ``practical'' compilers, including \texttt{gcc}, \texttt{GHC} and \texttt{SML/NJ}, are guilty of this sin, meaning that some programs that seem to be written in C, Haskell or Standard ML are actually written in tool-specific derivatives of those languages.
-Language-internal mechanisms do not lead to such fragmentation.} or other tool.
-For example, a researcher interested in providing regexp-related features (let us refer to these collectively as \texttt{R}) might design a new language with built-in support for regular expressions, perhaps basing it on an existing language (containing some general-purpose constructs, \texttt{G}).
-A different researcher developing a new language-based parallel programming abstraction or implementation strategy (collectively, \texttt{P}) might  take the same strategy.
-A third researcher, developing an alternative parallel programming abstraction (collectively, \texttt{Q}) might again do the same.
-This results in a collection of distinct systems as diagrammed in Figure \ref{approaches}a.%This latter method couples the semantics of the feature to the implementation details of a particular tool.
+When the semantics or implementation of a system must be extended to fully realize a new feature, as in the example above, providers typically take a \emph{language-external approach}, either by developing a new or derivative programming system (supported by \emph{language workbenches} \cite{erdweg2013state},  \emph{DSL frameworks} \cite{fowler2010domain} or \emph{compiler generators} \cite{abc}), or by extending an existing system by some mechanism that is not part of the language itself, such as an extension mechanism for a {particular} compiler\footnote{Compilers that modify, or allow modification of, the semantics of their base language, rather than simply permitting semantics-preserving optimizations, should be considered a pernicious means for creating new languages.
+Many ``practical'' compilers, including \texttt{gcc}, \texttt{GHC} and \texttt{SML/NJ}, are guilty of this, meaning that some programs that seem to be written in C, Haskell or Standard ML are actually written in tool-specific derivatives of these languages.
+Language-integrated mechanisms do not lead to such fragmentation.} or other tool.
+For example, a researcher interested in providing regular expression related features (let us refer to these collectively as \texttt{R}) might design a new system with built-in support for these, perhaps basing it on an existing system containing some general-purpose features (\texttt{G}).
+A different researcher developing a new language-based parallel programming abstraction or implementation strategy (\texttt{P}) might  take the same strategy.
+A third researcher, developing an alternative parallel programming abstraction (\texttt{Q}) might again do the same.
+This results in a collection of distinct systems as diagrammed in Figure \ref{approaches}a.
+Unfortunately, when providers of new features take language-external approaches like this, it causes  problems for clients related to \textbf{orthogonality} and \textbf{client compatibility}, described below.
+%This latter method couples the semantics of the feature to the implementation details of a particular tool.
 Because the use of one implementation entails a different semantics for the feature than another, the extended tool acts, \emph{de facto}, as a distinct system for our purposes.
 
-Unfortunately, when providers of new features take language-external approaches like this, it causes  problems for clients.
-Features cannot be adopted individually, but instead are only available coupled to a collection of other unrelated  features (e.g.
-the incidental features of the newly-designed language and its associated tools).
+\paragraph{Orthogonality} Features implemented by language-external means cannot be adopted individually, but instead are only available coupled to a fixed collection of other features.
 This makes adoption more costly when these incidental features are  not desirable or insufficiently developed, or when the features bundled with a different language or tool are simultaneously desirable.
-Evidence supports the prominence of this problem: developers prefer language-integrated parallel programming abstractions to library-based implementations if all else is equal \cite{cave2010comparing}, but library-based implementations are more widely adopted because ``parallel programming languages'' privilege only a few chosen  abstractions and implementation strategies.
-This is problematic because different parallel abstractions or implementation strategies are more appropriate in different situations \cite{Tasharofi:2013rc} and moreover,  parallel programming support is rarely the only concern relevant to client developers outside of a classroom setting.
+That is, one must either use the system containing features \texttt{G+R}, \texttt{G+P} or \texttt{G+Q}.
+There is no system containing \texttt{G}, \texttt{R}, \texttt{P} and \texttt{Q} in other combinations, and merging the systems containing each separately can be non-trivial because there is no requirement that providers of new features use a common mechanism.
+Even in cases where a common mechanism has been used, there are serious interference and safety issues, as we will discuss at length throughout this thesis (cf.
+Sec.
+\ref{background}, below).
+
+Recent evidence emphasizes the importance of this problem for driving research into practice: developers prefer language-integrated parallel programming abstractions to library-based implementations if all else is equal \cite{cave2010comparing}, but library-based implementations are more widely adopted because ``parallel programming languages'' privilege only a few chosen  abstractions and associated implementation strategies.
+This is problematic because different parallel programming abstractions or implementation strategies are more appropriate in different situations \cite{Tasharofi:2013rc}.
+Moreover,  parallel programming support is rarely the only concern relevant to client developers outside of a classroom setting.
 Regular expression support, for example, may be simultaneously desirable for processing large amounts of textual data in parallel, but using these features together in the same compilation unit when they have been implemented by language-external means is difficult or impossible.
-One must either use the system containing features \texttt{G+R} or \texttt{G+P}.
-There is no system containing both.
 %Similarly, a language and tools designed primarily to support regular expressions might make an interesting research project, but it would not be a suitable tool for writing large applications with more varied needs.
@@
 
-Different application concerns can sometimes be separated into different components, each written using a suitable language and tooling.
-This has been called the \emph{language-oriented approach} to software development \cite{journals/stp/Ward94}.
-Unfortunately, this is insufficient: the interface between components written in different languages remains an issue.
-The specialized constructs particular to one language (e.g.
-futures in a parallel programming language) cannot always be safely and naturally expressed in terms of those available in another (e.g.
-a general-purpose language), so building a program out of components written in a variety of different languages is difficult whenever these are exposed at interface boundaries.
+\paragraph{Client Compatibility} Even in cases where for each component of a system, there has been constructed a  system that is completely satisfactory, there remain problems at the interface between components.
+An interface that exposes the specialized constructs particular to one language (e.g.
+futures in a parallel programming language) cannot necessarily be safely and naturally consumed from another language (e.g.
+a general-purpose language).
 Tool support is also lost when calling into a different language.
-We call this fundamental issue the \emph{client compatibility problem}: clients of a certain collection of features cannot always interface with clients of a different collection of features in a safe, performant and natural manner.
+We call this fundamental issue the \emph{client compatibility problem}: code written by clients of a certain collection of features cannot always interface with code written by clients of a different collection  in a safe, performant and natural manner.
 
-One strategy often taken by proponents of the language-oriented approach to partially address the compatibility problem is to  target an established intermediate language, such as the Java Virtual Machine (JVM) bytecode, and support a superset of its constructs.
+One strategy often taken by proponents of a \emph{language-oriented approach} to software development \cite{journals/stp/Ward94} to partially address the client compatibility problem is to  target an established intermediate language, such as the Java Virtual Machine (JVM) bytecode, and use its constructs as a common language for communication between components written in different languages.
 Scala \cite{200464/IC} and F\# \cite{pickering2007foundations} are examples of prominent general-purpose languages that have taken this approach, and most DSL frameworks also rely on this strategy.
-As indicated in Figure \ref{approaches}a, this only enables true client compatibility in one direction.
-While calling into the common language becomes straightforward, calls in the other direction, or between the languages sharing the common target, are still restricted by the semantics of the common language.
+As indicated in Figure \ref{approaches}a, this only enables client compatibility in one direction.
+Calling into the common language becomes straightforward, but calling in the other direction, or between the languages sharing the common target, is not supported.
 
-If new languages only include constructs that can already be expressed safely and reasonably naturally in the common language, then this approach can work well.
-But many of the most innovative constructs found in modern languages (often, the features that justify their creation) are difficult to define in terms of existing constructs in ways that guarantee all necessary invariants are statically maintained and that do not require large amounts boilerplate code and run-time overhead.
+This approach can work well when new languages consist of constructs that can also be expressed safely and almost as naturally in the common language.
+But many of the most innovative constructs found in modern languages (often, those that justify their creation) are difficult to define in terms of existing constructs in ways that guarantee all necessary invariants are statically maintained and that do not require large amounts boilerplate code and run-time overhead.
 For example, the type system of F\# guarantees that null values cannot occur within F\# data structures, but maintaining this important invariant still requires run-time checks because the typing rules of F\# do not apply when F\# code is called from other languages on the Common Language Infrastructure (CLI) like C\#.
-The F\# type system also includes support for checking that units of measure are used correctly \cite{syme2012expert, kennedy1994dimension}, but this specialized invariant is left completely unchecked at language boundaries.
+The F\# type system also includes support for checking that units of measure are used correctly \cite{syme2012expert, kennedy1994dimension}, but this highly-specialized invariant is left completely unchecked at language boundaries.
 %In Scala, traits that have default method implementations are difficult to implement from Java or other JVM languages and the workaround can break if the trait is modified \cite{scalatraitinterop}. 
 In some cases, desirable features must be omitted entirely due to concerns about interoperability.
-F\#, for example, is based on Ocaml, but due to the need for bidirectional interoperability with CLI languages, it cannot support features like polymorphic variants, modules or functors \cite{ocaml-manual}.
+F\#, for example, aimed to retain source compatibility with Ocaml code, but due to the need for bidirectional interoperability with CLI languages, it does not support features like polymorphic variants, modules or functors \cite{ocaml-manual} because they have no straightforward analogs in the type system of the CLI.
 %\end{itemize}
 
-\subsection{Language-Integrated Approaches}
-We argue that, due to these problems, taking a language-external approach to realizing a new feature should be considered harmful and avoided whenever possible.
-The goal of the research being proposed here is to design \emph{language-integrated extension mechanisms} that give developers the ability to define in libraries new edit-time and compile-time features of the system that have previously been defined centrally\footnote{One might compare today's monolithic programming systems to  {centrally-planned} economies, whereas extensible\- systems more closely resemble modern market economies.} so that such language-external approaches are less frequently necessary.
-More specifically, we will show how control over aspects of \textbf{syntax}, \textbf{typechecking}, \textbf{elaboration} and \textbf{editor services} can be delegated to user-defined logic distributed in {libraries}, as illustrated in Figure \ref{approaches}b.
-Such libraries are called \emph{active libraries} because, rather than being passive clients of features already available in the system, they contain logic invoked by the system during development or compilation to provide new features  \cite{activelibraries}.
-Features implemented within active libraries can be imported as needed by the clients of libraries that rely on them, unlike features implemented by language-external means, seemingly avoiding the client compatibility problem.
+\subsection{Language-Integrated Approaches}\label{language-integrated-approaches}
+We argue that, due to these problems with orthogonality and client compatibility, taking a language-external approach to realizing a new feature should be considered harmful and avoided whenever possible.
+The goal of the research being proposed here is to design \emph{language-integrated extension mechanisms} that give providers the ability to define, within libraries, new features that have previously required central planning\footnote{One might compare today's programming systems to  {centrally-planned} economies, whereas extensible\- systems more closely resemble modern market economies.
+Our safety constraints serve a role analagous to market regulation.
+We leave further development of this analogy to the reader.} so that language-external approaches are less frequently necessary.
+More specifically, we will show how control over aspects of the \textbf{syntax}, \textbf{static and dynamic semantics} and \textbf{editor services} can be delegated to user-defined logic distributed in {libraries}, as illustrated in Figure \ref{approaches}b.
+Such libraries have been called \emph{active libraries}  \cite{activelibraries} because, rather than being passive clients of features already available in the system, they contain logic invoked by the system during development or compilation to provide new features.
+Features implemented within active libraries can be imported as needed, unlike features implemented by external means, seemingly avoiding the problems of orthogonality and client compatibility.
 
-We must proceed with caution, however: critical issues having to do with {safety} must be overcome before language-integrated extension mechanisms can be introduced into a programming system, because if too much control over such core aspects of the system is given  to developers, the system may become unreliable.
+We must proceed with caution, however: critical issues having to do with {safety} must be overcome before language-integrated extension mechanisms can be introduced into a programming system.
+If too much control over  these core aspects of the system is given  to developers, the system may become unreliable.
 %For example, an extension could weaken important metatheoretic guarantees previously provided by the system. 
@@
 Furthermore, even if extensions can be shown not to cause such problems in isolation, there may still be conflicts between extensions that could weaken their semantics, leading to subtle problems that only appear at link-time.
-For example, if two active libraries introduce the same syntactic form but back it with differing (valid) semantics, the issue would only manifest itself when both libraries were imported somewhere within the same compilation unit.
-These kinds of safety issues have plagued previous attempts to design language-integrated extensibility mechanisms.% To prevent them, our mechanisms will organize  extension logic around types to guarantee that extensions are both safe in isolation and also safely composable in any combination.
+For example, if two active libraries introduce the same syntactic form but back it with differing (but individually valid) semantics, the issue would only manifest itself when both libraries were imported somewhere within the same compilation unit.
+These kinds of safety issues have plagued previous attempts to design language-integrated extensibility mechanisms.
+We will briefly review some of these attempts below.% To prevent them, our mechanisms will organize  extension logic around types to guarantee that extensions are both safe in isolation and also safely composable in any combination.
 

### 2014-02-06 thesis-proposal/problem.tex
Updated missing references

@@
 Library providers thus need to compromise, typically by asking clients to provide regular expressions as strings and deferring parsing, typechecking and compilation to run-time.
-This introduces performance overhead and can lead to unanticipated run-time errors (as shown in \cite{spishak2012type}) and security vulnerabilities (due to injection attacks when user inputs are spliced into patterns, for example) \cite{string-stuff}.
+This introduces performance overhead and can lead to unanticipated run-time errors (as shown in \cite{spishak2012type}) and security vulnerabilities (due to injection attacks when user inputs are spliced into patterns, for example) \cite{Bravenboer:2007:PIA:1289971.1289975}.
 Useful tools for working with regular expressions are rarely integrated into editors, and even more rarely in a way that facilitates their discovery and use directly when the developer is manipulating regular expressions.
-Tools that must be discovered independently and accessed externally are used infrequently \cite{emersonStudy} and can be more awkward than necessary \cite{Campbell:2008:DRT:1636642.1636651, Omar:2012:ACC:2337223.2337324}, leading to lower productivity \cite{emersonStudy, Omar:2012:ACC:2337223.2337324}.\todo{complete references}
+Tools that must be discovered independently and accessed externally are used infrequently \cite{Murphy-Hill:2011:PIE:1958824.1958888} and can be more awkward than necessary \cite{Campbell:2008:DRT:1636642.1636651, Omar:2012:ACC:2337223.2337324}, leading to lower productivity.
 
@@
 
-When the semantics or implementation of a system must be extended to fully realize a new feature, as in the example above, providers typically take a \emph{language-external approach}, either by developing a new or derivative programming system (supported by \emph{language workbenches} \cite{erdweg2013state},  \emph{DSL frameworks} \cite{fowler2010domain} or \emph{compiler generators} \cite{abc}), or by extending an existing system by some mechanism that is not part of the language itself, such as an extension mechanism for a {particular} compiler\footnote{Compilers that modify, or allow modification of, the semantics of their base language, rather than simply permitting semantics-preserving optimizations, should be considered a pernicious means for creating new languages.
+When the semantics or implementation of a system must be extended to fully realize a new feature, as in the example above, providers typically take a \emph{language-external approach}, either by developing a new or derivative programming system (supported by \emph{language workbenches} \cite{erdweg2013state},  \emph{DSL frameworks} \cite{fowler2010domain} or \emph{compiler generators} \cite{brooker1963compiler}), or by extending an existing system by some mechanism that is not part of the language itself, such as an extension mechanism for a {particular} compiler\footnote{Compilers that modify, or allow modification of, the semantics of their base language, rather than simply permitting semantics-preserving optimizations, should be considered a pernicious means for creating new languages.
 Many ``practical'' compilers, including \texttt{gcc}, \texttt{GHC} and \texttt{SML/NJ}, are guilty of this, meaning that some programs that seem to be written in C, Haskell or Standard ML are actually written in tool-specific derivatives of these languages.

### 2014-02-11 thesis-proposal/acc.tex
Near-final proposal

--- 2013-09-10 thesis-proposal/acc.tex
@@
 % !TEX root = omar-thesis-proposal.tex
-\section{Active Code Completion}
+\section{Active Code Completion}\label{acc}
+A language's syntax and semantics influences the behavior of a variety of tools beyond the compiler.
+For example, software developers today make heavy use of the code completion support found in modern source code editors  \cite{murphy_how_2006}.
+Editors for object-oriented languages provide code completion in the form of a floating menu containing  contextually-relevant variables, fields, methods, types and other code snippets.
+By navigating and selecting from this menu, developers are able to avoid many common spelling and logic errors, eliminate unnecessary keystrokes and explore unfamiliar APIs without incurring the mental overhead associated with switching to an external documentation tool.
+% The way code completion menus brought documentation systems into the editor, active code completion interfaces bring external tools into the editor.
+
+%Several refinements and additions to the code completion menu have previously been suggested in the literature.
+These have focused on leveraging additional sources of information, such as databases of usage history \cite{robbes_how_2008}\cite{HouPletcher2011}, inheritance information \cite{HouPletcher2011}, API-specific information \cite{HouPletcher2011}\cite{Lee+2008}, partial abbreviations \cite{Han+2009}, examples extracted from code repositories \cite{bruch_learning_2009}\cite{Brandt+2010} and crowdsourced information \cite{mooty_calcite:_2010}\cite{SnipMatch}, to increase the relevance and sophistication of the featured menu items.
+As with the standard form of code completion, many of these sources of data can also be utilized via external tools (e.g.
+Calcite \cite{mooty_calcite:_2010} uses information that could already be accessed using the Jadeite \cite{conf/vl/StylosFYM09} tool).
+Empirical evidence presented in these studies, however, suggests that directly integrating these kinds of tools into the editor is particularly effective.
+For example, users of the Calcite tool completed 40\% more tasks in a lab study (unfortunately, a Jadeite control group was not included.)
+
+In all such systems, the code completion interface has remained primarily menu-based.
+When an item in the menu is selected, code is inserted immediately, without further input from the developer.
+These systems are difficult to extend: a fixed strategy determines the completions that are available, so library providers cannot directly specify new domain-specific or contextually-relevant logic or provide assistance beyond that which a menu can provide.
+In this paper we propose a technique called {\it active code completion} that eliminates these restrictions using active types.
+This  makes developing and integrating a broad array of highly-specialized code generation tools directly into the editor, via the familiar code completion command, significantly simpler.% This technique is motivated by the evidence discussed above and further evidence provided in this paper that developers prefer, and make more effective use of, tools that do not require leaving the immediate editing environment.
+
+In this work, we discuss active code completion in the context of object construction in Java because type-aware editors for Java are better than developed than those for other languages, because we wish to do user studies, and because Java already provides a way to associate metadata with classes.
+The techniques in this section apply equally well to type-aware editors for any language with a similar mechanism.
+
+For example, consider the following Java code fragment:
+
+\begin{lstlisting}
+  public Color getDefaultColor() {
+      return _
+\end{lstlisting}
+
+\begin{figure*}\label{color}
+\begin{center}
+\includegraphics[width=40pc]{color_palette.png}\end{center}
+\caption{(a) An example code completion palette associated with the \texttt{Color} class. (b) The source code generated by this palette.}
+\label{colorpalette}
+\end{figure*}
+
+If the developer invokes the code completion command at the indicated cursor position (\li{_}), the editor  looks  for a {\it palette definition} associated with the {\it type} of the expression being entered, which in this case is  \verb|Color|.
+If an associated palette is found, a menu item briefly describing this palette is added to the standard code completion menu.
+When selected, the corresponding palette is shown, replacing the standard code completion menu.
+Figure \ref{colorpalette}(a) gives an example of a simple palette that may be associated with the \verb|Color| class\footnote{A video demonstrating this process is available at \url{http://www.cs.cmu.edu/~NatProg/graphite.html}.}.
+
+The developer can interact with such palettes to provide parameters and other information related to their intent, and receive immediate feedback about the effect these choices will have on the behavior of the object being constructed.
+When this interaction is complete, the palette generates appropriate source code for insertion at the cursor.
+Figure \ref{colorpalette}(b) shows the inserted code after the user presses \textsc{enter}.
+
+We sought to address the following questions before designing and implementing our active code completion system:
+
+\begin{itemize}
+\item What {\it specific} use cases exist for this form of active code completion in a professional development setting? 
+\item What {\it general} criteria are common to types that would and would not benefit from an associated palette?
+\item What are some relevant usability and design criteria for palettes designed to address such use cases?
+\item What capabilities must the underlying active code completion system provide to enable these use cases and user interface designs?
+\end{itemize}
+
+To help us answer these questions, we conducted a survey of 473 professional developers.
+Their responses, along with information gathered from informal interviews and code corpus analyses, revealed a number of non-trivial functional requirements for palette interfaces as well as the underlying active code completion architecture.
+Participants also suggested a large number of use cases, demonstrating the broad applicability of this technique.
+We organized these into several broad categories.
+Next, we developed Graphite, an Eclipse plug-in that implements the active code completion architecture for the Java programming language, allowing Java library developers to associate custom palettes with their own classes.
+We describe several design choices that we made to satisfy the requirements discovered in our preliminary investigations and examine necessary trade-offs.
+Finally, we conducted a pilot lab study with a more complex palette, implemented using Graphite, that assists developers as they write regular expressions.
+The study provides specific evidence in support of the broader claim that highly-specialized tools that are integrated directly with the editing environment are  useful.
+We conclude that active code completion systems like Graphite are useful because they make developing, deploying and discovering such tools fundamentally simpler.
+
+The primary concerns relevant to this thesis are:
+\begin{itemize}
+\item The palette mechanism should not be tied to a specific editor implementation.
+We achieve this by using a URL-based scheme for referring to palettes, which are implemented as webpages, which can be embedded into any editor using standard techniques for embedding browsers into GUIs.
+\item The palette mechanism should not be able to arbitrary access the surrounding source code (for privacy reasons, as identified by survey participants).
+By using a browser, and only allowing access to highlighted strings, we avoid this problem.
+\item We provide a mechanism by which users can associate palettes with types externally.
+This could cause conflicts with palettes that the type defines itself.
+To resolve this, we give users a choice whenever such situations occur by inserting all relevant palettes into the code completion menu.
+\end{itemize}
+
+\subsection{Remaining Tasks and Timeline}
+This work has been published at ICSE 2012 \cite{omar-graphite12}, and we do not plan on extending it further in this thesis.
+The main tasks have to do with relating it to the mechanisms in the previous sections, so that it generalizes beyond Java.
+We plan to do this when writing the thesis.
+There are also some pieces of related work that have been published since this paper was accepted that we need to review.

### 2014-02-11 thesis-proposal/parsing.tex
Near-final proposal

@@
 stringstyle=\ttfamily,
-numbers=left, numberstyle=\scriptsize\color{gray}\ttfamily, language=ML,moredelim=[il][\sffamily]{?},mathescape=true,showspaces=false,showstringspaces=false,xleftmargin=15pt,escapechar=@, morekeywords=[1]{tyfam,opfam,let,fn,val,def,casetype,objtype,metadata,of,*,datatype,new,valAST},deletekeywords={for,double,in},classoffset=0,belowskip=\smallskipamount,
+numbers=left, numberstyle=\scriptsize\color{gray}\ttfamily, language=ML,moredelim=[il][\sffamily]{?},mathescape=true,showspaces=false,showstringspaces=false,xleftmargin=15pt, morekeywords=[1]{tyfam,opfam,let,fn,val,def,casetype,objtype,metadata,of,*,datatype,new,valAST},deletekeywords={for,double,in},classoffset=0,belowskip=\smallskipamount,
 moredelim=**[is][\color{cyan}]{SSTR}{ESTR},
@@
 \end{lstlisting}
-By encoding lists in this way, we can immediately reason about them by structural induction, introduce them by naming the appropriate case and providing the associated data, if any, and examine them by pattern matching.
+By encoding lists in this way, we can immediately reason about them by structural induction and examine them by pattern matching.
 %The programmer only needs to provide an encoding of the structure of lists; the semantics and implementation are filled in by the general-purpose abstraction mechanism.
@@
 \subsection{Implementing a TSL}
-Portions of the implementation of the TSL for \li{HTML} are shown on lines 8-17 of Fig.
+Portions of the implementation of the TSL for \li{HTML} are shown on lines 8-15 of Fig.
 \ref{f-htmltype}.
@@
 
-Each production is followed, in an indented block, by a Wyvern function that generates a value given $n$ arguments, which will bound to the values recursively generated by each of the $n$ non-terminals it contains, ordered left-to-right.
-For the starting non-terminal, always called \li{start}, this function should return a value of type \li{Exp}.
+Each production is followed, in an indented block, by a Wyvern function that generates a value given the values recursively generated by each of the $n$ non-terminals in contains, ordered left-to-right.
+For the starting non-terminal, always called \li{start}, this function must return a value of type \li{Exp}.
+User-defined non-terminals might have a different type associated with them (not shown).
 Here, we show how to generate an AST using general-purpose notation for \li{Exp} (lines 13-15) as well as a more natural \emph{quasiquote} style (lines 11 and 18).
-Quasiquotes are expressions that are not evaluated, but rather reified into their corresponding syntax tree.
+Quasiquotes are expressions that are not evaluated, but rather reified into syntax trees.
 We observe that quasiquotes too fall into the pattern of ``specialized notation associated with a type'' -- quasiquotes for expressions, types and identifiers are simply TSLs associated with \li{Exp}, \li{Type} and \li{ID} (Fig.
@@
 Again, interpolation is type safe and structured, rather than based on string interpolation.
-This is accomplished by placing the interpolated value in a place in the generated AST where its type is known -- on line 11 of Fig.
+
+We have now seen several examples of TSLs that support interpolation.
+The question then arises: what type should the interpolated Wyvern expression be expected to have?
+This is determined by placing the interpolated value in a place in the generated AST where its type is known -- on line 11 of Fig.
 \ref{f-htmltype} it is known to be \li{HTML} and on line 13 it is known to be \li{CSS} by the declaration of \li{HTML}, and on line 15, it is known to be \li{HTML} by the use of an explicit ascription.
@@
 \begin{enumerate}
-\item We provide a more complete layout-sensitive concrete syntax.
-We show how it can be written without the need for a context-sensitive lexer or parser and give a full specification for the layout-delimited literal form introduced by a forward reference, \li{~}, as well as other forms of forward-referenced blocks.
+\item We provide a complete layout-sensitive concrete syntax.
+We show how it can be written without the need for a context-sensitive lexer or parser using Adams' formalism and provide a full specification for the layout-delimited literal form introduced by a forward reference, \li{~}, as well as other forms of forward-referenced blocks (for the forms \li{new} and \li{case(e)}, in particular).
 %\item We detail the general mechanism for associating metadata with a type.
@@
 \subsection{Remaining Tasks and Timeline}
-The following tasks remain to be completed. I have scheduled this for the period of March 2nd-25th, before the OOPSLA submission deadline.
+The paper we have submitted for ECOOP 2014 has received quite positive reviews, so we anticipate that it will be accepted (notifications are sent on Mar.
+7th).
+Other than final revisions on the paper text, due on May 12th, the following tasks remain to be completed.
+We will largely do these tasks after the OOPSLA deadline in late March.
 \begin{enumerate}
-\item In our current top-level grammar, expressions within parenthesis must still obey layout constraints.
+%\item In our current top-level grammar, expressions within parenthesis must still obey layout constraints.
 This is not strictly necessary for preventing ambiguities, and it is quite convenient to not require this (as in Python, for example).
@@
 It may be possible to prevent shadowing by making it impossible for variables to be introduced into interpolated Wyvern expressions (so that function application is the only way to pass data from a TSL to Wyvern code, as in the \li{Parser} TSL we have shown).
-\item The corpus analysis we conducted was preliminary. We must perform this in a more complete and rigorous manner (collaborative work with Darya Kurilova).
+%\item The corpus analysis we conducted was preliminary. We must perform this in a more complete and rigorous manner (collaborative work with Darya Kurilova).
 %\item The current implementation of Wyvern does not include all of these mechanisms as described.

### 2014-02-11 thesis-proposal/timeline.tex
Near-final proposal

@@
 \section{Timeline}
+To summarize, we plan on completing the work in this thesis per the following schedule:
 \begin{itemize}
-\item @$\lambda$ - ICFP - Mar. 1
-\item Wyvern - ECOOP or OOPSLA - late March
-\item Ace - PLDI or OOPSLA - late March
-\item Then write everything up.
+\item The work on @$\lambda$ will be submitted on Mar. 1 to ICFP. 
+\item The work on Ace will be submitted on Mar. 25 to OOPSLA.
+\item The work on Wyvern will, pending likely final acceptance, be written up in final form by May 12 for ECOOP.
+\item The work on Graphite has been completed and published at ICSE.
 \end{itemize}

### 2014-02-17 thesis-proposal/acc.tex
Updated proposal

@@
 
-In this work, we discuss active code completion in the context of object construction in Java because type-aware editors for Java are better than developed than those for other languages, because we wish to do user studies, and because Java already provides a way to associate metadata with classes.
+In this work, we discuss active code completion in the context of object construction in Java because type-aware editors for Java are better developed than those for other languages, because we wish to do empirical studies, and because Java already provides a way to associate metadata with classes.
 The techniques in this section apply equally well to type-aware editors for any language with a similar mechanism.

### 2014-02-17 thesis-proposal/background.tex
Updated proposal

@@
 A prominent example in the literature is Blitz++, a library that uses C++ template metaprogramming to optimize compound operations on vectors and matrices by eliminating intermediate allocations \cite{veldhuizen2000blitz++}.
-Although this and several other interesting optimizations have been shown possible by this technique, its expressiveness is fundamentally limited because template expansion allows for only the substitution of compile-time constants into pre-written code, and template metaprograms are notoriously difficult to read, write, debug and reason about (see discussion in \cite{Robison:2001:IEC:376656.376751}).
+Although this and several other interesting optimizations are possible by this technique, its expressiveness is fundamentally limited because template expansion allows for only the substitution of compile-time constants into pre-written code, and template metaprograms are notoriously difficult to read, write, debug and reason about (see discussion in \cite{Robison:2001:IEC:376656.376751}).
 %They are, however, composable, because templates are associated with a single function or class, and  the only metatheoretic issue is that template expansion can be non-terminating.

### 2014-02-17 thesis-proposal/parsing.tex
Updated proposal

@@
 basicstyle=\ttfamily\fontsize{8pt}{1em}\selectfont, 
-commentstyle=\itshape\rmfamily, 
+commentstyle=\itshape\ttfamily, 
 stringstyle=\ttfamily,
@@
 moredelim=**[is][\color{Periwinkle}]{SGRM}{EGRM},
-moredelim=**[is][\color{YellowGreen}]{SID}{EID}
+moredelim=**[is][\color{YellowGreen}]{SID}{EID},
+moredelim=**[is][\color{Sepia}]{SUS}{EUS}
 }
@@
 Among the most common strategies in these situations is to simply use a string representation that is parsed at run-time.
-Developers across language paradigms frequently write examples like those above as:
+Developers across language paradigms frequently write examples like the above as:
 \begin{lstlisting}
@@
 To allow us to focus on the essence of our proposal, the variant of Wyvern we will describe in this thesis is simpler than the variant previously described: it is purely functional (there are no effects or mutable state) and it does not enforce a uniform access principle for objects (fields can be accessed directly).
-Objects are thus essentially labeled product types with simple methods (functions that require a self-reference).
+Objects are thus essentially recursive labeled product types with simple methods (functions that require a self-reference).
 We also add recursive labeled sum types, which we call \emph{case types}, that are quite similar to datatypes in ML.
@@
 \ref{f-example} with an example showing several different TSLs being used to define a fragment of a web application showing search results from a database.
-We will go through this example below to develop intuitions about TSLs in Wyvern.
 Note that for clarity of presentation, we color each character  according to the TSL it is governed by.
@@
 This is an {object type} declaring several fields representing the components of a URL: its protocol, domain name, port, path and so on.
-We could have created a value of type \li{URL} using general-purpose notation\todo{fix comment formatting}\todo{new color}:
+We could have created a value of type \li{URL} using general-purpose notation:
 \begin{lstlisting}
 let imageBase : URL = new
-  val protocol : URLString = $\text{"}$SSTRhttpESTR$\texttt{"}$
-  val subdomain : URLString = $\texttt{"}$SSTRimagesESTR$\texttt{"}$
+  val protocol : URLString = $\text{"}$SUShttpEUS$\texttt{"}$
+  val subdomain : URLString = $\texttt{"}$SUSimagesEUS$\texttt{"}$
   (* ... *)
@@
 
-Each production is followed, in an indented block, by a Wyvern function that generates a value given the values recursively generated by each of the $n$ non-terminals in contains, ordered left-to-right.
+Each production is followed, in an indented block, by a Wyvern function that generates a value given the values recursively generated by each of the $n$ non-terminals it contains, ordered left-to-right.
 For the starting non-terminal, always called \li{start}, this function must return a value of type \li{Exp}.
@@
   ... 
-  | FromTS of Exp * Exp
+  | FromTS of Tokenstream
   | Literal of TokenStream
@@
 \subsection{Formalization}
-A formal and more detailed description can be found in our paper draft\footnote{\url{https://github.com/wyvernlang/docs/tree/master/ecoop14}}. In particular:
+A formal and more detailed description can be found in our paper draft.\footnote{\url{https://github.com/wyvernlang/docs/blob/master/ecoop14/ecoop14.pdf?raw=true}} In particular:
 \begin{enumerate}
@@
 We observe that \emph{grammars} and \emph{quasiquotes} can both be seen as TSLs for parsers and ASTs respectively and discuss how to implement them as such.
-\item We formalize the static semantics, including the literal parsing logic, of TSL Wyvern by combining a bidirectional type system (in the style of Lovas and Pfenning \cite{lovasPfenning}) with an elaboration semantics (in the style of Harper and Stone \cite{harperStone}).
+\item We formalize the static semantics, including the literal parsing logic, of TSL Wyvern by combining a bidirectional type system (in the style of Lovas and Pfenning \cite{Lovas08abidirectional}) with an elaboration semantics (in a style similar to Harper and Stone \cite{Harper00atype-theoretic}).
 By distinguishing locations where an expression synthesizes a type from locations where an expression is being analyzed against a known type, we can precisely state where generic literals can and cannot appear and how parsing is delegated to a TSL.
@@
 We formalize this hygiene mechanism by bifurcating the context in our static semantics.
-\item We provide several examples of TSLs throughout the paper, but to examine how broadly applicable the technique is, we conduct a simple corpus analysis, finding that string languages are used ubiquitously in existing Java code  (collaborative work with Darya Kurilova\todo{and Josh?}).
+\item We provide several examples of TSLs throughout the paper, but to examine how broadly applicable the technique is, we conduct a simple corpus analysis, finding that string languages are used ubiquitously in existing Java code  (collaborative work with Darya Kurilova).
 \end{enumerate}
@@
 \subsection{Remaining Tasks and Timeline}
-The paper we have submitted for ECOOP 2014 has received quite positive reviews, so we anticipate that it will be accepted (notifications are sent on Mar.
+The paper we have submitted to ECOOP 2014 has received quite positive reviews, so we anticipate that it will be accepted (notifications are sent on Mar.
 7th).
-Other than final revisions on the paper text, due on May 12th, the following tasks remain to be completed.
-We will largely do these tasks after the OOPSLA deadline in late March.
+If accepted, final revisions on the paper text are due on May 12th and so we will largely do these tasks after the OOPSLA deadline in late March.
+If not accepted, we will aim to resubmit to OOPSLA.
 \begin{enumerate}
@@
 \item We must write down the full formal semantics (including rules that we omitted for concision in the paper draft) in a technical report, as well as provide more complete proofs of the metatheory.
+This might require slightly restricting the recursion in the rule for literals, so that induction is well-founded.
 \item Our current static semantics does not support mutually recursive type declarations, but this is necessary for our prelude to typecheck, so we will add support for this.
-\item We must further consider aspects of hygiene.
-In particular, we do not yet have a clean mechanism for preventing unintentional variable \emph{shadowing}, only unintentional variable \emph{capture}.
-It may be possible to prevent shadowing by making it impossible for variables to be introduced into interpolated Wyvern expressions (so that function application is the only way to pass data from a TSL to Wyvern code, as in the \li{Parser} TSL we have shown).
+In the paper, the formalism is meant to be expository, so we will likely leave this additional complexity for the tech report.
+\item We must further consider aspects of hygiene:
+\begin{enumerate}
+\item We do not yet have a clean mechanism for preventing unintentional variable \emph{shadowing}, only unintentional variable \emph{capture}.
+It may be possible to prevent shadowing by making it impossible for variables to be introduced into interpolated Wyvern expressions directly (so that function application is the only way to pass data from a TSL to Wyvern code, as in the \li{Parser} TSL above).
+\item In macro systems like Scheme's, free variables in generated ASTs refer to their values within the macro definition, rather than where they are inserted.
+To support this, we have introduced the \li{valAST} operator.
+Using it is a bit verbose and awkward -- every free variable in a quasiquote must be inside \li{valAST}.
+We believe we can insert \li{valAST} automatically from within the definition of the TSL for \li{Exp} using a very simple extension to the interpolation mechanism that allows interpolation of single variables, rather than whole expressions, and plan to write up how this works.
+\end{enumerate}
 %\item The corpus analysis we conducted was preliminary. We must perform this in a more complete and rigorous manner (collaborative work with Darya Kurilova).

### 2014-02-17 thesis-proposal/problem.tex
Updated proposal

@@
 In their place, we will develop \emph{language-integrated extensibility mechanisms} that decentralize control over several core aspects of the programming system.
-By organizing new features around {types} and constraining them appropriately, we aim to show that these mechanisms can guarantee safety and non-interference of extensions while also being highly expressive.
+By organizing new features around {types} and constraining them appropriately, we aim to show that these mechanisms can guarantee safety and non-interference of extensions while remaining highly expressive.
 %But taking a \emph{language-internal approach} to implementing a feature is the most practical.
@@
 For example, a researcher interested in providing regular expression related features (let us refer to these collectively as \texttt{R}) might design a new system with built-in support for these, perhaps basing it on an existing system containing some general-purpose features (\texttt{G}).
-A different researcher developing a new language-based parallel programming abstraction or implementation strategy (\texttt{P}) might  take the same strategy.
+A different researcher developing a new language-based parallel programming abstraction or implementation strategy (\texttt{P}) might  take the same approach.
 A third researcher, developing an alternative parallel programming abstraction (\texttt{Q}) might again do the same.
 This results in a collection of distinct systems as diagrammed in Figure \ref{approaches}a.
-Unfortunately, when providers of new features take language-external approaches like this, it causes  problems for clients related to \textbf{orthogonality} and \textbf{client compatibility}, described below.
+Unfortunately, when providers of new features take language-external approaches like this, it causes  problems for clients related to \textbf{orthogonality} and \textbf{client compatibility}.
 %This latter method couples the semantics of the feature to the implementation details of a particular tool.
@@
 
-Recent evidence emphasizes the importance of this problem for driving research into practice: developers prefer language-integrated parallel programming abstractions to library-based implementations if all else is equal \cite{cave2010comparing}, but library-based implementations are more widely adopted because ``parallel programming languages'' privilege only a few chosen  abstractions and associated implementation strategies.
-This is problematic because different parallel programming abstractions or implementation strategies are more appropriate in different situations \cite{Tasharofi:2013rc}.
+Recent evidence indicates that this is one of the major barriers preventing research from being driven into practice: developers prefer language-integrated parallel programming abstractions to library-based implementations if all else is equal \cite{cave2010comparing}, but library-based implementations are more widely adopted because ``parallel programming languages'' privilege only a few chosen  abstractions and associated implementation strategies.
+This is problematic because different parallel programming abstractions or implementation strategies are seen as more appropriate in different situations \cite{Tasharofi:2013rc}.
 Moreover,  parallel programming support is rarely the only concern relevant to client developers outside of a classroom setting.
@@
 
-\paragraph{Client Compatibility} Even in cases where for each component of a system, there has been constructed a  system that is completely satisfactory, there remain problems at the interface between components.
+\paragraph{Client Compatibility} Even in cases where for each component of a system there is a  system that is completely satisfactory, there remain problems at the interface between components.
 An interface that exposes the specialized constructs particular to one language (e.g.
@@
 In some cases, desirable features must be omitted entirely due to concerns about interoperability.
-F\#, for example, aimed to retain source compatibility with Ocaml code, but due to the need for bidirectional interoperability with CLI languages, it does not support features like polymorphic variants, modules or functors \cite{ocaml-manual} because they have no straightforward analogs in the type system of the CLI.
+F\#, for example, aimed to retain source compatibility with Ocaml code, but due to the need for bidirectional interoperability with CLI languages, it does not support features like polymorphic variants, modules or functors \cite{ocaml-manual} because they have no apparent analogs in the type system of the CLI.
 %\end{itemize}
@@
 We argue that, due to these problems with orthogonality and client compatibility, taking a language-external approach to realizing a new feature should be considered harmful and avoided whenever possible.
-The goal of the research being proposed here is to design \emph{language-integrated extension mechanisms} that give providers the ability to define, within libraries, new features that have previously required central planning\footnote{One might compare today's programming systems to  {centrally-planned} economies, whereas extensible\- systems more closely resemble modern market economies.
+The goal of the research being proposed here is to design \emph{language-integrated extension mechanisms} that give providers the ability to define, within libraries, new features that have previously required central planning
+%\footnote{One might compare today's programming systems to  {centrally-planned} economies, whereas extensible\- systems more closely resemble modern market economies.
 Our safety constraints serve a role analagous to market regulation.
-We leave further development of this analogy to the reader.} so that language-external approaches are less frequently necessary.
+We leave further development of this analogy to the reader.}
+so that language-external approaches are less frequently necessary.
 More specifically, we will show how control over aspects of the \textbf{syntax}, \textbf{static and dynamic semantics} and \textbf{editor services} can be delegated to user-defined logic distributed in {libraries}, as illustrated in Figure \ref{approaches}b.

### 2014-02-17 att-icfp14/att-icfp14.tex
Updated ICFP14

@@
 
-\title{Active Typechecking and Translation: A Safe Language-Internal Extension Mechanism}
-
-\authorinfo{~}{~}{~}
-%\authorinfo{Cyrus Omar\and Jonathan Aldrich}
- %          {School of Computer Science\\
-  %          Carnegie Mellon University}
-   %        {\{comar, jonathan.aldrich\}@cs.cmu.edu}
+\title{@$\lambda$: Conservatively Extending a Type System From Within}
+
+\authorinfo{Cyrus Omar\and Jonathan Aldrich}
+           {Carnegie Mellon University}
+            {\{comar, aldrich\}@cs.cmu.edu}
 
@@
 \begin{abstract}
-Researchers and domain experts often propose new language primitives as extensions to the semantics of an existing language. 
-But today's statically-typed languages are {monolithic}: they do not expose language-internal mechanisms for implementing the static and dynamic semantics of new primitive types and their associated operators directly, so these experts must instead create new standalone languages.
-This causes problems for potential users because building applications from components written in many different languages can be both unsafe and unnatural. 
-An {internally-extensible} language could  address these issues, but designing a mechanism that is expressive while maintaining safety remains a challenge. 
-Extensions must be modularly verified, their use in any combination must not weaken the metatheoretic properties of the language, nor can they interfere with one another.
-We introduce a mechanism called active type-checking and translation (AT\&T) that aims to directly address these issues while remaining highly expressive. 
-AT\&T leverages type-level computation, typed compilation techniques and a form of type abstraction to enable library-based implementations of a variety of primitives over a flexible  grammar in a safe manner.
+Researchers often extend an existing language with new type and operator constructors. 
+But because this is not generally possible from within the language, researchers must implement extensions as distinct forks of the language.
+This limits their utility: building applications where different components are written in languages with different type systems is unsafe and unnatural. 
+An {internally-extensible} type system could help address these issues, but designing an extension mechanism that is both expressive and reliable is a challenge: extensions must not be permitted to compromise key metatheoretic properties of the language, nor should they be allowed to interfere with one another, in any combination.
+This latter property, \emph{conservativity}, is not guaranteed by prior mechanisms.
+
+We introduce @$\lambda$, a simply-typed lambda calculus with simply-kinded type-level computation where the  semantics of new type and operator constructors can be defined in a functional style from within.
+Its semantics are structured as an extensible variant of the Harper-Stone elaboration semantics used to specify Standard ML.
+We incorporate typed compilation techniques and a form of type abstraction that together allow us to prove type safety, decidability and conservativity.
+We intend @$\lambda$ to serve as a minimal foundation for future work on extension mechanisms for typed programming languages, but it is already quite expressive.
+We begin by showing how a more natural concrete syntax can be introduced by a type-directed desugaring to @$\lambda$.
+Then, we show how a number of abstractions that must be built into other languages can be implemented orthogonally, as libraries, within @$\lambda$.
+We conclude with a discussion of its present limitations.
 \end{abstract}
@@
 %type-level computation, typed compilation
-\section{Motivation}
-When designing and implementing a new abstraction, experts typically begin by attempting to define new constructs in terms of existing language constructs.
-This approach is often effective because modern {general-purpose} abstraction mechanisms, like inductive datatypes and object systems, are highly expressive.  
-For example, the Delite framework leverages Scala's powerful general-purpose mechanisms to enable a number of useful \emph{embedded domain-specific languages} \cite{delite}.
-Unfortunately, there remain some situations of interest where general-purpose abstractions fall short. 
-For instance, it is difficult to adequately encode advanced type systems in terms of the simpler rules that govern general-purpose abstractions (e.g.
+\section{Introduction}
+Typed programming languages are typically organized around a collection of indexed type and operator constructors.
+The simply-typed lambda calculus (STLC), for example, provides a single type constructor, $\rightarrow$, indexed by a pair of types, and two operator constructors, $\lambda$, indexed by a type, and $\opvar{ap}$, which can be thought of as being indexed trivially.
+We can extend it with universally quantified and recursive types by adding the type constructors $\forall$ and $\mu$, each indexed by a type binder (a type-level term that may refer to a free type variable), and their associated operator constructors.
+It is tempting to stop here: the language is now universal \todo{right terminology?} (i.e.
+Turing complete) and the dynamic semantics of operators associated with other useful types, including sums and products, can be defined by a method analagous to Church encodings in the untyped lambda calculus \cite{reynolds}.
+This is of clear theoretical and pedagogical interest.
+But this calculus is plainly impractical as a programming language.
+Reynolds, striking a cautionary tone reminiscent of Perlis in his description of the ``Turing tarpit'' \cite{perlis}, put it succinctly \cite{reynolds}:
+\begin{quote}
+To say that any reasonable function can be expressed by some program is not to say that it can be expressed by the most reasonable program.
+It is clear that the language requires a novel programming style.
+\end{quote}
+
+A simple  language suffices when only the dynamic semantics of a program are being considered but continuing to add type and operator constructors (collectively, \emph{constructs}) to a language can make reasoning about programs simpler and more precise and support a more natural programming style.
+Consistent with this view, typed programming languages generally expose more constructs externally than their compilers work with internally.
+A functional language like ML, for example, might expose $n$-ary product types (tuples), labeled product types (records), recursive labeled sum types (datatypes) and forms of type abstraction (polymorphism and abstract data types).
+A compiler for the language might then translate these to an intermediate  language (IL) with only simple products and sums for further compilation to machine code.
+Correctly implementing the dynamic semantics of the language is the primary concern during this phase, so a simple IL that decreases the number of cases that must be considered when implementing and verifying the compiler is a wise choice.
+
+It is again tempting to stop here.
+Indeed, languages like ML are often referred to as ``general-purpose languages'' and have been used successfully for a range of computing tasks.
+The constructs they provide certainly occupy ``sweet spots'' in the design space, and many abstractions can be encoded using these constructs in a completely satisfactory manner.
+However, they admit no new ``universality'' theorems  stronger than those for the simple calculus described above, and situations continue to arise in both research and practice where new constructs are desirable:
+
+\begin{enumerate}
+\item General-purpose constructs continue to evolve.
+There are many  variations on record types, for example: structurally-typed records, or records with forms of dynamic dispatch.
+We will show an example of records with  prototypic dispatch in Sec.
+\ref{prototypes}, and argue that they cannot be safely and naturally defined in terms of standard functional constructs.
+Sum types also admit variants, as we will discuss in the next section.
+
+\item Specialized type systems that enforce stronger invariants than general-purpose abstraction mechanisms can enforce are often proposed by researchers.
+One need not take more than a cursory glance through the literature to discover type systems for parallel programming \cite{a,b}, concurrent programming \cite{cml,a,b}, distributed programming \cite{tom7}, dataflow programming \cite{reactiveml}, authenticated data structures \cite{popl13}, information flow \cite{walker00, smith2001}, databases \cite{db}, aliasing \cite{naden12}, effects \cite{effects},  network protocols \cite{sekar99}, regular expressions \cite{regex}, units of measure \cite{keneddy} and many others.
+\item Interoperability layers that allow programmers to safely and naturally interact with libraries written in a different language require enforcing the type system of that language within the calling language.
+Using library-based interoperability layers today can lead to safety issues when this is not possible.
+%\item General-purpose abstraction mechanisms are typically implemented uniformly.
+For example, inductive datatypes are typically implemented using tagged pointers to a payload.
+In some cases (e.g.
+natural numbers) it may be desirable to implement an abstraction in an alternative manner (e.g.
+as an integer) so that certain operations perform better.
+Parallel abstractions are of particular note, as they can benefit substantially from specialized implementation strategies on specialized hardware (e.g.
+a GPU).
+Note that if control over internal implementation details were the only relevant use case, abstract datatypes might be appropriate.
+\end{enumerate}
+
+Today, complete solutions to these problems generally take the form of a \emph{fork} of an existing general-purpose language (discussed further in the next section).
+This is unsatisfying because then these solutions are  \emph{non-orthogonal}: a programmer can choose either a language supporting a reasonable approach to parallel programming  or one that builds in support for working with and statically reasoning about regular expressions, but there is not necessarily a language that supports both.
+Even if a separation of such concerns can be achieved, such that each component of an application can be written using a specialized language (as advocated by proponents of the \emph{language-oriented approach} to software development \cite{language-oriented}), there will be an \emph{interoperability} problem at the interface between components, as described in item 3 above.
+
+  %We can specify the static semantics of the STLC as follows, using a uniform abstract syntax where type and operator indices are written in braces to emphasize this way of thinking about its structure:
+%\begin{figure}[h]
+%\small
+%\vspace{-8pt}
+%\begin{mathpar}
+%\inferrule[var]{ }{
+%	\jet{\eCtxX{\eCtx}{x}{\tau}}{x}{\tau}
+%}
+%
+%\inferrule[arrow-I]{
+%	\jet{\eCtxX{\eCtx}{x}{\tau}}{e}{\tau'}	
+%}{
+%	\jetX{\lambda[\tau]({x}.e)}{\fvar{Arrow}[(\tau, \tau')]}
+%}
+%
+%\inferrule[arrow-E]{
+%	\jetX{e_1}{\fvar{Arrow}[(\tau, \tau')]}\\
+%	\jetX{e_2}{\tau}
+%}{
+%	\jetX{\texttt{ap}[()](e_1; e_2)}{\tau'}
+%}
+%\end{mathpar}
+%\vspace{-8pt}
+%\caption{Static semantics of the simply typed lambda calculus.}
+%\end{figure}
+
+\todo{everything below is todo}Rather than specifying a complete language, a researcher might simply specify the new type and operator constructors and
+
+%\begin{figure}[h]
+%\small
+%\vspace{-8pt}
+%\begin{mathpar}
+%\inferrule[nat-I1]{ }{
+%	\jetX{\texttt{z}[()]()}{\fvar{Nat}[()]}
+%}
+%
+%\inferrule[nat-I2]{
+%	\jetX{e}{\fvar{Nat}[()]}
+%}{
+%	\jetX{\texttt{s}[()](e)}{\fvar{Nat}[()]}
+%}
+%
+%\inferrule[nat-E]{
+%	\jetX{e_1}{\fvar{Nat}[()]}\\
+%	\jetX{e_2}{\tau}\\
+%	\jet{\eCtxX{\eCtxX{\eCtx}{x}{\fvar{Nat}[()]}}{y}{\tau}}{e_3}{\tau}
+%}{
+%	\jetX{\texttt{natrec}[()](e_1; e_2; x.y.e_3)}{\tau}
+%}
+%\end{mathpar}
+%\vspace{-8pt}
+%\caption{Static semantics of natural numbers.}\label{nat-statics}
+%\vspace{-15pt}
+%\end{figure}
+%\begin{figure}[h]
+%\small
+%\begin{mathpar}
+%\inferrule[Prod-I]{
+%	\jetX{e_1}{\tau_1}\\
+%	\jetX{e_2}{\tau_2}
+%}{
+%	\jetX{\texttt{pair}[()](e_1; e_2)}{\fvar{Prod}[(\tau_1, \tau_2)]}
+%}
+%
+%\inferrule[Prod-E1]{
+%	\jetX{e}{\fvar{Prod}[(\tau_1, \tau_2)]}
+%}{
+%	\jetX{\texttt{prl}[()](e)}{\tau_1}
+%}
+%
+%\inferrule[Prod-E2]{
+%	\jetX{e}{\fvar{Prod}[(\tau_1, \tau_2)]}
+%}{
+%	\jetX{\texttt{prr}[()](e)}{\tau_1}
+%}
+%\end{mathpar}
+%\vspace{-8pt}
+%\caption{Static semantics of products}\label{pair-statics}
+%\end{figure}
+
+%creating a new language.
+If this is not practical, the best one can attempt to do is encode the new types in terms of existing types (by a Church encoding, for example).
+This is generally unsatisfactory --
+
+%Languages implemented using these common patterns are central planning by a language designer or design committee.
+
+%Researchers or domain experts who cannot work around such limitations must develop new standalone languages.
+In our simple scenario, we may simply copy our implementation of G\"odel's T or even edit it directly (a pernicious technique for implementing a new language where the prior one is overwritten).
+In a more complex scenario, we may instead employ a tool like a compiler generator or DSL framework \cite{fowler2010domain} that can generate a standalone implementation from declarative specifications of language constructs.
+Some of these tools allow you to package and reuse these specifications (with the important caveat that not all combinations of constructs are valid and free of conflicts, an important modularity issue that we will return to several times in this paper).
+%
+%The increasing sophistication and ease-of-use of these tools have led many to suggest a {\it language-oriented approach} \cite{journals/stp/Ward94} to software development where different components of an application are written in different languages.
+Unfortunately, this leads to problems at language boundaries: a library's external interface must only use constructs that can reasonably be expressed in \emph{all possible calling languages}.
+This can restrict domain-specific languages by, for example, precluding constructs that rely on statically-checked invariants stronger than those their underlying representation in a common target language normally supports.
+At best, constructs like these can be exposed by generating a wrapper where run-time checks have been inserted to guarantee necessary invariants.
+This compromises both verifiability and performance and requires the development of an interoperability layer for every DSL.
+Moreover, library clients must work with verbose and unnatural ``glue code'' when interfacing across languages, defeating the primary purpose of high-level programming languages: hiding the low-level details from the end-users of abstractions.
+We diagram this fundamental \emph{compatibility problem} in Figure \ref{approaches}(a).
+%\begin{figure*}
+%\begin{center}
+%\includegraphics[scale=0.5]{approaches.pdf}
+%\end{center}
+%\vspace{-20px}
+%\caption{\small (a) With a language-oriented approach, novel constructs are packaged into separate languages.
+Users can only safely and naturally call into languages consisting of common constructs (often only the common target language, such as C or Java bytecode).
+(b) With a language-internal extensibility approach, there is one system providing a common internal language, where additional primitive constructs that strictly strengthen its static guarantees or perform specialized code generation are specified and distributed within libraries.
+\label{approaches}}
+%\end{figure*}
+
+%As a result, domain-specific languages and new general-purpose abstractions alike have experienced relatively slow adoption in practice.
+%
+%Porting large codebases to new languages is difficult, and the dominant programming languages innovate slowly, so programming language.
+%
+%More specifically, such languages are neither \emph{internally extensible} because the language itself exposes only natural numbers and functions to its users, nor are they \emph{externally extensible} because no new behaviors can be added to the language's  implementation in a separate module from the one containing the initial implementation.
+
+%This is the essence of a monolithic language implementation: it is impossible for anyone to modularly extend languages defined in this way.
+
+An extensible programming language could address these problems by providing a language-integrated mechanism for introducing new type and operator constructors and implementing their associated static and dynamic semantics directly.
+%Library developers need only consider which abstractions are most appropriate for their domain, without also considering whether these constructs can be exposed using abstractions appropriate to the domains of client code.
+Clients can simply import any necessary constructs when using a library that relies on them, preserving safety and ease-of-use without the use of  wrappers and glue code.
+We show this competing approach in Figure \ref{approaches}(b).
+%Researchers and domain experts thus gain the ability to distribute new ideas for evaluation to a broader development community without requiring the approval of maintainers of mainstream languages, large-scale porting of code or explicit interoperability layers.
+But, as mentioned in Section \ref{language-integrated-approaches}, some significant challenges must be addressed before such a mechanism can be relied upon.
+The desire for expressiveness must be balanced against  concerns about maintaining various safety properties in the presence of arbitrary combinations of user-defined  extensions to the language's core semantics.
+The mechanism must ensure that desirable \emph{metatheoretic properties} (e.g.
+type safety, decidability) of the language are maintained by extensions.
+Because multiple independently developed extensions might be used within one program, the mechanism must further guarantee their \emph{non-interference}.
+These are the issues we seek to address in this work.
+
+%When designing and implementing a new abstraction, experts typically begin by attempting to define new constructs in terms of existing language constructs.
+%This approach is often effective because modern {general-purpose} abstraction mechanisms, like inductive datatypes and object systems, are highly expressive.  
+%For example, the Delite framework leverages Scala's powerful general-purpose mechanisms to enable a number of useful \emph{embedded domain-specific languages} \cite{delite}.
+%Unfortunately, there remain some situations of interest where general-purpose abstractions fall short. 
+%For instance, it is difficult to adequately encode advanced type systems in terms of the simpler rules that govern general-purpose abstractions (e.g.
 reasoning about units of measure requires built-in language support in F\# \cite{conf/cefp/Kennedy09}).
-Even if a full encoding is possible, it may not be useful if it is overly verbose or unnatural, or if the error messages are overly abstract.
+%Even if a full encoding is possible, it may not be useful if it is overly verbose or unnatural, or if the error messages are overly abstract.
 For example, regular expressions encoded using inductive datatypes are quite verbose, so most functional languages support them via strings, which is less safe.
-Finally, general-purpose abstractions are implemented in a uniform manner, meaning domain-specific heuristics cannot be applied to eliminate overhead or perform optimizations, and implementations designed for typical application workloads may not be satisfactory in parts of a program where performance is a key criteria, such as when targeting heterogeneous hardware platforms (e.g.
+%Finally, general-purpose abstractions are implemented in a uniform manner, meaning domain-specific heuristics cannot be applied to eliminate overhead or perform optimizations, and implementations designed for typical application workloads may not be satisfactory in parts of a program where performance is a key criteria, such as when targeting heterogeneous hardware platforms (e.g.
 programmable GPUs) and distributed computing resources.
-
-% are at times impossible or impractical.
+%
+%% are at times impossible or impractical.
 In our example of adding products or sums to Godel's T, although Church encodings are possible \cite{pfpl}, they require a reasonable level of creativity\footnote{Anecdotally, Church encodings are among the more difficult-to-explain topics covered in our undergraduate programming languages course.}.
@@
 Scala?}
-
-%creating a new language.
+%
+%%creating a new language.
 If this is not practical, the best one can attempt to do is encode the new types in terms of existing types (by a Church encoding, for example).
 This is generally unsatisfactory --
-
-%Languages implemented using these common patterns are central planning by a language designer or design committee.
-
-Researchers or domain experts who run into situations like these, where more direct control over a language's semantics and implementation are needed, have little choice today but to realize new abstractions by creating a new language in some way.
+%
+%%Languages implemented using these common patterns are central planning by a language designer or design committee. 
+%
+%Researchers or domain experts who run into situations like these, where more direct control over a language's semantics and implementation are needed, have little choice today but to realize new abstractions by creating a new language in some way.
 They might develop a new standalone language from scratch, modify an implementation of an existing language, or use tools like compiler generators, DSL frameworks and language workbenches \cite{fowler2010domain}.
-%In our simple scenario, we may simply fork our implementation of Godel's T or even edit it directly (a pernicious technique for implementing a new language where the prior one is overwritten). 
-%In a more complex scenario, we may instead employ a tool like a compiler generator or DSL framework \cite{fowler2010domain} that can generate a standalone implementation from declarative specifications of language constructs.
+%%In our simple scenario, we may simply fork our implementation of Godel's T or even edit it directly (a pernicious technique for implementing a new language where the prior one is overwritten). 
+%%In a more complex scenario, we may instead employ a tool like a compiler generator or DSL framework \cite{fowler2010domain} that can generate a standalone implementation from declarative specifications of language constructs.
 Some of these tools allow you to package and reuse these specifications (with the important caveat that not all combinations of constructs are valid and free of conflicts, an important modularity issue that we will return to several times in this paper).
-The increasing sophistication and ease-of-use of these tools have led to calls for a {\it language-oriented approach} to software development, where different components of an application are written in different specialized languages \cite{journals/stp/Ward94}.
+%The increasing sophistication and ease-of-use of these tools have led to calls for a {\it language-oriented approach} to software development, where different components of an application are written in different specialized languages \cite{journals/stp/Ward94}.
 Indeed, a number of software ecosystems are now designed explicitly to support many different languages, both general-purpose and domain-specific, atop a common intermediate language.
 The Java virtual machine (JVM), the Common Language Infrastructure (CLI) and LLVM are prominent examples of such ecosystems.
-
-Unfortunately, this leads to a critical problem at language boundaries: a library's external interface must only use constructs that can reasonably be expressed in \emph{all possible client languages}.
+%
+%Unfortunately, this leads to a critical problem at language boundaries: a library's external interface must only use constructs that can reasonably be expressed in \emph{all possible client languages}.
 This discourages languages from including constructs that rely on statically-checked invariants stronger than those supported by their underlying implementation in the common intermediate language.
@@
 %Forrequires the development of an interoperability layer for every pair of DSLs.
-Moreover, this approach exposes the internals of an implementation to clients, making the abstraction awkward to work with and causing code breakage when implementation details change.
+%Moreover, this approach exposes the internals of an implementation to clients, making the abstraction awkward to work with and causing code breakage when implementation details change.
 This defeats a primary purpose of high-level programming languages: hiding low-level details from clients of an abstraction.
 We diagram this fundamental \emph{interoperability problem} in Figure \ref{approaches}(a).
-%As an example, F\#'s type system prevents \lstinline{null} values from occurring within data structures, but because it's type system is not available when calling into F\# code from another language, like C\#, run-time null checks must still be included in the implementation.
-\begin{figure*}[t]
-\vspace{-15px}
-\includegraphics[scale=0.415]{approaches.pdf}
-\caption{(a) With the language-oriented approach, different primitive abstractions are packaged into separate languages that extend and target a common intermediate language (e.g.
+%%As an example, F\#'s type system prevents \lstinline{null} values from occurring within data structures, but because it's type system is not available when calling into F\# code from another language, like C\#, run-time null checks must still be included in the implementation.
+%%\begin{figure*}[t]
+%%\vspace{-15px}
+%%\includegraphics[scale=0.415]{approaches.pdf}
+%%\caption{(a) With the language-oriented approach, different primitive abstractions are packaged into separate languages that extend and target a common intermediate language (e.g.
 JVM bytecode).
@@
 Clients can import and use these abstractions directly whenever needed.}
-\label{approaches}
-\end{figure*}
-
-%As a result, domain-specific languages and new general-purpose abstractions alike have experienced relatively slow adoption in practice.
-%
-%Porting large codebases to new languages is difficult, and the dominant programming languages innovate slowly, so programming language.
-%
-%More specifically, such languages are neither \emph{internally extensible} because the language itself exposes only natural numbers and functions to its users, nor are they \emph{externally extensible} because no new behaviors can be added to the language's  implementation in a separate module from the one containing the initial implementation.
-
-%This is the essence of a monolithic language implementation: it is impossible for anyone to modularly extend languages defined in this way.
-
-%Programming languages are typically designed around a monolithic collection of primitive type families and operators.
+%%\label{approaches}
+%%\end{figure*}
+%
+%%As a result, domain-specific languages and new general-purpose abstractions alike have experienced relatively slow adoption in practice.
+%%
+%%Porting large codebases to new languages is difficult, and the dominant programming languages innovate slowly, so programming language.
+%%
+%%More specifically, such languages are neither \emph{internally extensible} because the language itself exposes only natural numbers and functions to its users, nor are they \emph{externally extensible} because no new behaviors can be added to the language's  implementation in a separate module from the one containing the initial implementation.
+%
+%%This is the essence of a monolithic language implementation: it is impossible for anyone to modularly extend languages defined in this way. 
+%
+%%Programming languages are typically designed around a monolithic collection of primitive type families and operators.
 Consider, as a simple example, Godel's T \cite{pfpl}, a typed lambda calculus with recursion on primitive natural numbers\todo{add statics to Appendix A}.
@@
 That is, Godel's T is not \emph{internally extensible}.
-
-\emph{Internally-extensible programming languages} promise to avoid these problems by providing researchers and domain experts with a mechanism for implementing the semantics of new primitive constructs directly within libraries.
-%Developers of libraries need only determine whether they are appropriate for their domain, without also considering whether these constructs can be exposed in terms of abstractions appropriate to client code.
-As a result, clients can granularly import any necessary primitive constructs when using code that relies on them, and thereby achieve full safety, ease-of-use and performance.
+%
+%\emph{Internally-extensible programming languages} promise to avoid these problems by providing researchers and domain experts with a mechanism for implementing the semantics of new primitive constructs directly within libraries.
+%%Developers of libraries need only determine whether they are appropriate for their domain, without also considering whether these constructs can be exposed in terms of abstractions appropriate to client code.
+%As a result, clients can granularly import any necessary primitive constructs when using code that relies on them, and thereby achieve full safety, ease-of-use and performance.
 Providers of components thus need only consider whether primitives that they use are appropriate for their domain, without also considering whether their code might be used in a context where these primitives are not otherwise appropriate.
@@
 We adopt this terminology and diagram this competing approach in Figure \ref{approaches}(b).
-
-%Researchers and domain experts thus gain the ability to distribute new ideas for evaluation to a broader development community without requiring the approval of maintainers of mainstream languages, large-scale porting of code or explicit interoperability layers.
-
-For a language-internal extension mechanism to be feasible, however, it must achieve expressiveness while also ensuring that extensions cannot compromise the safety properties of the language and its tools, nor interfere with one another.
+%
+%%Researchers and domain experts thus gain the ability to distribute new ideas for evaluation to a broader development community without requiring the approval of maintainers of mainstream languages, large-scale porting of code or explicit interoperability layers.
+%
+%For a language-internal extension mechanism to be feasible, however, it must achieve expressiveness while also ensuring that extensions cannot compromise the safety properties of the language and its tools, nor interfere with one another.
 That is, extensions cannot simply be permitted to add arbitrary logic to the type system or compiler, because this would make it possible to break  type safety, decidability or adequacy theorems that are critical to the operation of the language, the compiler or other extensions.
 We review some previous attempts at language extensibility, and highlight how they do not adequately achieve both safety and expressiveness, in Section \ref{related-work}.
-%{\color{red} transition here} Correctness properties of an extension itself should be modularly verifiable, so that its users can rely on it for verifying and compiling their own code.
+%%{\color{red} transition here} Correctness properties of an extension itself should be modularly verifiable, so that its users can rely on it for verifying and compiling their own code.
 The mechanism must also ensure that desirable metatheoretic properties and global safety guarantees of the language cannot be weakened by extensions.
 And with multiple independently-developed extensions used at once, the mechanism must further guarantee that they cannot interfere with one another.
-
-In this paper, we introduce a language-internal extensibility mechanism called \emph{active typechecking and translation} (AT\&T) that allows developers to introduce and implement the logic governing new primitive type and operator families from within libraries.
-We argue that this can be accomplished by enriching the type-level language, rather than introducing a separate metalanguage into the system. 
-To make this proposal concrete, we begin by introducing a simple core calculus, called \atlam~(for the ``actively-typed lambda calculus''), in Section \ref{atlam}. 
-This calculus uses type-level computation of higher kind, along with techniques borrowed from the typed compilation literature and a form of type abstraction that ensures that the implementation details of an extension are not externally visible to guarantee the safety of the language, the decidability of typechecking and compilation and non-interference of extensions, as we outline in Section \ref{safety}.
-In Section \ref{examples}, we suggest that despite these constraints, this mechanism is expressive enough to admit, within libraries, a number of general-purpose and domain-specific abstractions that normally require built-in language support.
-
+%
+%In this paper, we introduce a language-internal extensibility mechanism called \emph{active typechecking and translation} (AT\&T) that allows developers to introduce and implement the logic governing new primitive type and operator families from within libraries.
+%We argue that this can be accomplished by enriching the type-level language, rather than introducing a separate metalanguage into the system. 
+%To make this proposal concrete, we begin by introducing a simple core calculus, called \atlam~(for the ``actively-typed lambda calculus''), in Section \ref{atlam}. 
+%This calculus uses type-level computation of higher kind, along with techniques borrowed from the typed compilation literature and a form of type abstraction that ensures that the implementation details of an extension are not externally visible to guarantee the safety of the language, the decidability of typechecking and compilation and non-interference of extensions, as we outline in Section \ref{safety}.
+%In Section \ref{examples}, we suggest that despite these constraints, this mechanism is expressive enough to admit, within libraries, a number of general-purpose and domain-specific abstractions that normally require built-in language support.
+%
 %Our core calculus uses a uniform abstract syntax for primitive operators to simplify our presentation and analysis, but this syntax is too verbose to be practical.
@@
 \section{From Extensible Compilers to Extensible Languages}\label{evolution}
-To understand the genesis of our internal extension mechanism, it is helpful to begin by considering why most implementations of programming languages cannot even be  externally extended. 
-Let us consider, as a simple example, an implementation of G\"odel's T, a typed lambda calculus with recursion on primitive natural numbers (see Appendix). 
-A compiler for this language written using a functional language will invariably represent the primitive type families and operators using {closed} inductive datatypes. 
-For example, a simple implementation in Standard ML may be based around these datatypes:
-\begin{lstlisting}
-  datatype Type = Nat | Arrow of Type * Type
-  datatype Exp = Var of var 
-               | Lam of var * Type * Exp | Ap of Exp * Exp 
-               | Z | S of Exp | Natrec of Exp * Exp * Exp
-\end{lstlisting}
-
-The logic governing typechecking and translation to a suitable intermediate language (for subsequent optimization and compilation by some back-end) will proceed by exhaustive case analysis over the constructors of \lstinline{Exp}.
-
-In an object-oriented implementation of Godel's T, we might instead encode types and operators as subclasses of abstract classes \lstinline{Type} and \lstinline{Exp}.
-Typechecking and translation will proceed by the ubiquitous \emph{visitor pattern}  by dispatching against a fixed collection of {known} subclasses of \lstinline{Exp}.
-
-In either case, we encounter the same basic issue: there is no way to modularly add new primitive type families and operators and implement their associated typechecking and translation logic. 
-%This issue is related to the widely-discussed \emph{expression problem} (in a restricted sense -- we do not consider adding new functions beyond typechecking and translation here, only adding logic to these) \cite{wadler-expression}.
-
-A number of language mechanisms have been proposed that allow new cases to be added to datatypes and the functions that operate over them in a modular manner. 
-In functional languages, we might use \emph{open datatypes}.
-For example, if we wish to extend G\"odel's T with product types and we have written our compiler in a language supporting open inductive datatypes, it might be possible to add new cases like this:
-\begin{lstlisting}
-  newcase Prod of Type * Type extends Type
-  newcase Pair of Exp * Exp extends Exp    (* Intro *)
-  newcase PrL of Exp extends Exp           (* Elim Left *)
-  newcase PrR of Exp extends Exp           (* Elim Right *)
-\end{lstlisting}
-
-The logic for functionality like typechecking and translation could then be implemented for only these new cases.
-For example, the \lstinline{typeof} function that assigns a type to an expression could be extended like so:
-\begin{lstlisting}
-  typeof PrL(e) = case typeof e of 
-      Prod(t1, _) => t1 
-    | _ => raise TypeError("<appropriate error message>")
-\end{lstlisting}
-
-If we allowed users to define new modules containing definitions like these and link them into our compiler, we will have succeeded in creating an externally-extensible compiler, albeit one where safety is not guaranteed (we will return to this point shortly).
-We have not, however, created an extensible programming language, for two reasons.
-First, compiler extensions are distributed and activated separately from libraries, so dependencies become more difficult to manage.
-Second, other compilers for the same language will not necessarily support the same extensions.
-If our newly-introduced constructs are exposed at a library's  interface boundary, clients using different compilers face the same problems with interoperability that those using different languages face.
-That is, {extending a language by extending a single compiler for it is morally equivalent to creating a new language}.
-Several prominent language ecosystems today are in a state where a prominent compiler has introduced or enabled the introduction of extensions that many libraries have come to rely on, including the Glasgow Haskell Compiler, SML/NJ and the GNU compilers for C and C++.
-
-A more appropriate and useful place for extensions like this is directly within libraries, alongside abstractions that can be adequately implemented in terms of existing primitive abstractions.
-To enable this, the language must allow for the introduction new primitive type families, like \lstinline{Prod}, operators, like \lstinline{Pair}, \lstinline{PrL} and \lstinline{PrR}, and associated typechecking and translation logic.
-When encountering these new operators in expressions, the compiler must effectively  hand control over typechecking and translation to the appropriate user-defined logic.
-Because this mechanism is {language-internal}, all compilers must support it to satisfy the language specification.
-
-Statically-typed languages typically make a distinction between \emph{expressions}, which describe run-time computations, and type-level constructs like types, type aliases and datatype declarations.
-The design described above suggests we may now need to add another layer to our language, an {extension language}, where extensions can be declared and implemented.
-In fact, we will show that \textbf{the most natural place for type system extensions is within the type-level language}.
-The intuition is that extensions to a statically-typed language's semantics will need to manipulate types as values at compile-time.
-Many languages already allow users to write type-level functions for various reasons, effectively supporting this notion of types as values at compile-time (see Sec.
-\ref{related-work} for examples).
-The type-level language is often constrained by its own type system (where the types of type-level values are called \emph{kinds} for clarity) that prevents type-level functions from causing problems during compilation.
-This is precisely the structure that a distinct extension layer would have, and so it is quite natural to unify the two, as we will show in this work.
-
-\section{\atlam}\label{atlam}
 \begin{figure}[t]
@@
 $$\begin{array}{rccl}	
-\textbf{programs} & \rho & ::= & \pfam{\familyDf}{\progsort} \pipe \pdef{t}{\kappa}{\tau}{\progsort} \pipe e\\
+\textbf{programs} & \rho & ::= & \pfam{\familyDf}{\progsort}  \\
+ & & \pipe &  \pdef{t}{\kappa}{\tau}{\progsort} \pipe e\\
 \text{primitive ops}		&	\theta	&	::= &	\tops{op}{\kappaidx}{i}{a}{\tau} \pipe 
@@
 \\
-\textbf{expressions} 				&	e	&	::=	&	\evar{x} \pipe 
+\textbf{external terms} 				&	e	&	::=	&	\evar{x} \pipe 
 														\elam{\evar{x}}{\tau}{e} \pipe 
@@
 														\tlam{t}{\kappa}{\tau} \pipe 
-														\tapp{\tau_1}{\tau_2} \pipe
+														\tapp{\tau_1}{\tau_2}\pipe
+														\iintlit \pipe \iop{\tau_1}{\tau_2} \pipe \tstr{str} \\
+									    & &  \pipe & 
 														\tnil{\kappa} \pipe \tcons{\tau_1}{\tau_2} \pipe 
@@
 
-\text{type-level data}	 			& 		& \pipe	& 	 \iintlit \pipe \iop{\tau_1}{\tau_2} \pipe \tstr{str} \pipe \tunit \pipe 
+	 			& 		& \pipe	& 	 \tunit \pipe 
 														\tpair{\tau_{1}}{\tau_{2}} \pipe 
@@
 														\\
-\text{structural equality}  & & \pipe & 					\tifeq{\tau_{1}}{\tau_{2}}{\kappa}{\tau_{3}}{\tau_{4}} 
-														\\
-
-\text{types} 						& 		& \pipe	& 	\ttypestd \pipe \tfamcase{\tau}{Fam}{x}{\tau_1}{\tau_2}\\
+\text{equality}  & & \pipe & 					\tifeq{\tau_{1}}{\tau_{2}}{\kappa}{\tau_{3}}{\tau_{4}} 
+														\\
+
+\text{types} 						& 		& \pipe	& 	\ttypestd \\
+ & & \pipe &  \tfamcase{\tau}{Tycon}{x}{\tau_1}{\tau_2}\\
 						%				& & \pipe & \tfamcase{\tau}{Fam}{x}{\tau_1}{\tau_2}\\
 
-\text{denotations} 				& 		 & 	\pipe	&	\tden{\tauiterm}{\tautype} \pipe \terr \pipe \tdencase{\tau}{x}{t}{\tau_1}{\tau_2}\\
+\text{denotations} 				& 		 & 	\pipe	&	\tden{\tauiterm}{\tautype} \pipe \terr \\
+ & & \pipe &  \tdencase{\tau}{x}{t}{\tau_1}{\tau_2}\\
  %& & \pipe & 
@@
 												\kunit \pipe 
-												\kpair{\kappa_{1}}{\kappa_{2}} \pipe 
+												\kpair{\kappa_{1}}{\kappa_{2}} \\
+	& & \pipe &  
 												\kTypeBlur \pipe \kDen \pipe 
@@
 												\iapp{\gamma_{1}}{\gamma_{2}} \pipe
-												\ifix{\evar{f}}{\sigma}{\gamma} \pipe
+												\ifix{\evar{f}}{\sigma}{\gamma} \\
+									& & \pipe & 
 												\ipair{\gamma_{1}}{\gamma_{2}} \pipe 
@@
 \end{array}$$
-\vspace{-10pt}
+%\vspace{-10pt}
 \caption{\small Syntax of \atlam.
@@
 \end{figure}
-In this section, we will develop a core calculus, called @$\lambda$ for the ``actively-typed lambda calculus'', by way of a semantics and a simple example, and discuss how it addresses the safety concerns that arise.
-\subsection{Overview}
-The grammar of \atlam~is shown in Figure \ref{grammar}. 
-A program, $\rho$, consists of a series of declarations followed by an expression.
-Declarations can be either bindings of type-level terms to type-level variables using \textsf{def} or a primitive type family declared using \textsf{family}, which can contain implementations of one or more operator families, $\theta$.
-Expressions, $e$, can be either variables, lambdas, or applications of operators.
-%, and are ultimately given meaning by translation to a typed internal language, with terms $\gamma$ and types $\sigma$. 
-%This language has been chosen, for simplicity, to be a variant of Plotkin's PCF with primitive integers and products, but in practice would include other constructs consistent with its role as a high-level intermediate language.
-
-The language is structured as a simply-typed lambda calculus with simply-kinded type-level computation. Kinds, $\kappa$, classify type-level terms, $\tau$. 
-Types are type-level values of kind $\star$ (following System $F_{\omega}$) and classify expressions.
-The type-level language also includes other kinds of terms: type-level functions, lists (required by our mechanism), integers, strings and products for the sake of our examples (see Sec.
-\ref{safety} for a discussion on other acceptable kinds of type-level data) and constructs for developing extensions -- denotations and reified internal terms and types -- which we will discuss in the sections below.
-
-All expressions are given meaning by translation to a typed internal language.
-This language has been chosen, for simplicity, to be a variant of Plotkin's PCF with primitive integers and products, but in practice would include other constructs consistent with its role as a high-level intermediate language.
-The grammar of internal terms, $\gamma$, and internal types, $\sigma$, also includes special forms containing type-level terms; these are used for developing extensions and during compilation and will be erased before compilation ends.
 \begin{figure}[t]
 \small
-%\begin{lstlisting}
-%family Nat of 1 ~ _.\/(int) {
-%	z(i:1, a. case len a =_Z 0 then [[Nat<()> ==> 0]] else err);
-%	s(i:1, a. get_last_arg a err (\a1. 
-%	  dencase a1 of [[t1 ==> v1]] => 
-%	    if t1 =_* Nat<()> then [[Nat<()> ==> ^(v1) + 1]] else err
-%	  ow err))
-%}
-%\end{lstlisting}
 \begin{flalign}
@@
 \end{flalign}
-\vspace{-15pt}
 \caption{G\"odel's T in \atlam, used to calculate 2+2.
@@
 \end{figure}
+
+To understand the genesis of our internal extension mechanism, it is helpful to begin by considering why most implementations of programming languages cannot even be  externally extended. 
+Let us consider, as a simple example, an implementation of G\"odel's T, a typed lambda calculus with recursion on primitive natural numbers (see Appendix). 
+A compiler for this language written using a functional language will invariably represent the primitive type families and operators using {closed} inductive datatypes. 
+For example, a simple implementation in Standard ML may be based around these datatypes:
+\begin{lstlisting}
+  datatype Type = Nat | Arrow of Type * Type
+  datatype Exp = Var of var 
+               | Lam of var * Type * Exp | Ap of Exp * Exp 
+               | Z | S of Exp | Natrec of Exp * Exp * Exp
+\end{lstlisting}
+
+The logic governing typechecking and translation to a suitable intermediate language (for subsequent optimization and compilation by some back-end) will proceed by exhaustive case analysis over the constructors of \lstinline{Exp}.
+
+In an object-oriented implementation of Godel's T, we might instead encode types and operators as subclasses of abstract classes \lstinline{Type} and \lstinline{Exp}.
+Typechecking and translation will proceed by the ubiquitous \emph{visitor pattern}  by dispatching against a fixed collection of {known} subclasses of \lstinline{Exp}.
+
+In either case, we encounter the same basic issue: there is no way to modularly add new primitive type families and operators and implement their associated typechecking and translation logic. 
+%This issue is related to the widely-discussed \emph{expression problem} (in a restricted sense -- we do not consider adding new functions beyond typechecking and translation here, only adding logic to these) \cite{wadler-expression}.
+
+A number of language mechanisms have been proposed that allow new cases to be added to datatypes and the functions that operate over them in a modular manner. 
+In functional languages, we might use \emph{open datatypes}.
+For example, if we wish to extend G\"odel's T with product types and we have written our compiler in a language supporting open inductive datatypes, it might be possible to add new cases like this:
+\begin{lstlisting}
+  newcase Prod of Type * Type extends Type
+  newcase Pair of Exp * Exp extends Exp    (* Intro *)
+  newcase PrL of Exp extends Exp           (* Elim Left *)
+  newcase PrR of Exp extends Exp           (* Elim Right *)
+\end{lstlisting}
+
+The logic for functionality like typechecking and translation could then be implemented for only these new cases.
+For example, the \lstinline{typeof} function that assigns a type to an expression could be extended like so:
+\begin{lstlisting}
+  typeof PrL(e) = case typeof e of 
+      Prod(t1, _) => t1 
+    | _ => raise TypeError("<appropriate error message>")
+\end{lstlisting}
+
+If we allowed users to define new modules containing definitions like these and link them into our compiler, we will have succeeded in creating an externally-extensible compiler, albeit one where safety is not guaranteed (we will return to this point shortly).
+We have not, however, created an extensible programming language, for two reasons.
+First, compiler extensions are distributed and activated separately from libraries, so dependencies become more difficult to manage.
+Second, other compilers for the same language will not necessarily support the same extensions.
+If our newly-introduced constructs are exposed at a library's  interface boundary, clients using different compilers face the same problems with interoperability that those using different languages face.
+That is, {extending a language by extending a single compiler for it is morally equivalent to creating a new language}.
+Several prominent language ecosystems today are in a state where a prominent compiler has introduced or enabled the introduction of extensions that many libraries have come to rely on, including the Glasgow Haskell Compiler, SML/NJ and the GNU compilers for C and C++.
+
+A more appropriate and useful place for extensions like this is directly within libraries, alongside abstractions that can be adequately implemented in terms of existing primitive abstractions.
+To enable this, the language must allow for the introduction new primitive type families, like \lstinline{Prod}, operators, like \lstinline{Pair}, \lstinline{PrL} and \lstinline{PrR}, and associated typechecking and translation logic.
+When encountering these new operators in expressions, the compiler must effectively  hand control over typechecking and translation to the appropriate user-defined logic.
+Because this mechanism is {language-internal}, all compilers must support it to satisfy the language specification.
+
+Statically-typed languages typically make a distinction between \emph{expressions}, which describe run-time computations, and type-level constructs like types, type aliases and datatype declarations.
+The design described above suggests we may now need to add another layer to our language, an {extension language}, where extensions can be declared and implemented.
+In fact, we will show that \textbf{the most natural place for type system extensions is within the type-level language}.
+The intuition is that extensions to a statically-typed language's semantics will need to manipulate types as values at compile-time.
+Many languages already allow users to write type-level functions for various reasons, effectively supporting this notion of types as values at compile-time (see Sec.
+\ref{related-work} for examples).
+The type-level language is often constrained by its own type system (where the types of type-level values are called \emph{kinds} for clarity) that prevents type-level functions from causing problems during compilation.
+This is precisely the structure that a distinct extension layer would have, and so it is quite natural to unify the two, as we will show in this work.
+
+\section{\atlam}\label{atlam}
+In this section, we will develop a core calculus, called @$\lambda$ for the ``actively-typed lambda calculus'', by way of a semantics and a simple example, and discuss how it addresses the safety concerns that arise.
+\subsection{Overview}
+The grammar of \atlam~is shown in Figure \ref{grammar}. 
+A program, $\rho$, consists of a series of declarations followed by an expression.
+Declarations can be either bindings of type-level terms to type-level variables using \textsf{def} or a primitive type family declared using \textsf{family}, which can contain implementations of one or more operator families, $\theta$.
+Expressions, $e$, can be either variables, lambdas, or applications of operators.
+%, and are ultimately given meaning by translation to a typed internal language, with terms $\gamma$ and types $\sigma$. 
+%This language has been chosen, for simplicity, to be a variant of Plotkin's PCF with primitive integers and products, but in practice would include other constructs consistent with its role as a high-level intermediate language.
+
+The language is structured as a simply-typed lambda calculus with simply-kinded type-level computation. Kinds, $\kappa$, classify type-level terms, $\tau$. 
+Types are type-level values of kind $\star$ (following System $F_{\omega}$) and classify expressions.
+The type-level language also includes other kinds of terms: type-level functions, lists (required by our mechanism), integers, strings and products for the sake of our examples (see Sec.
+\ref{safety} for a discussion on other acceptable kinds of type-level data) and constructs for developing extensions -- denotations and reified internal terms and types -- which we will discuss in the sections below.
+
+All expressions are given meaning by translation to a typed internal language.
+This language has been chosen, for simplicity, to be a variant of Plotkin's PCF with primitive integers and products, but in practice would include other constructs consistent with its role as a high-level intermediate language.
+The grammar of internal terms, $\gamma$, and internal types, $\sigma$, also includes special forms containing type-level terms; these are used for developing extensions and during compilation and will be erased before compilation ends.
 \subsection{Example: G\"odel's T as an Active Type Family}
@@
 \end{mathpar}
-\vspace{-10pt}
 \caption{\small Central Compilation Judgement of \atlam.}
@@
 \end{mathpar}
-\vspace{-10pt}
 \caption{\small Active typechecking and translation of programs. Evaluation semantics for type-level terms are given in Fig. \ref{tleval}.}
@@
 }
-~~~~~~
+
 \inferrule[den-intro-err]{ }{
@@
 \inferrule[p-eq]{
-	\kEq{\kappa_1}\\
+	\kEq{\kappa_1}\\\\
 	\kEq{\kappa_2}
@@
 
-\begin{theorem}[Representational Consistency]
-If $\progOK{ }{\Sigma}{e}$ and $\ecompiles{ }{\fvalCtx}{e}{\tau}{\gabs}$ and $\eraseX{\gabs}{\gamma}$ and 	$\ddbar{\Xi_0}{\fvalCtx}{\trepof{\tau}}{\sabs}$ and $\eraseX{\sabs}{\sigma}$ then $\checkRC{ }{\Xi_0}{\fvalCtx}{\gamma}{\sigma}$.
-\end{theorem}
+%\begin{theorem}[Representational Consistency]
+%If $\progOK{ }{\Sigma}{e}$ and $\ecompiles{ }{\fvalCtx}{e}{\tau}{\gabs}$ and $\eraseX{\gabs}{\gamma}$ and 	$\ddbar{\Xi_0}{\fvalCtx}{\trepof{\tau}}{\sabs}$ and $\eraseX{\sabs}{\sigma}$ then $\checkRC{ }{\Xi_0}{\fvalCtx}{\gamma}{\sigma}$.
+%\end{theorem}
 
@@
 
-\begin{theorem}[Kind Safety and Termination]
-If $\vdash_{\Sigma} \tau : \kappa$ then $\tEvalX{\tau}{\tau'}$ such that $\vdash_{\Sigma_0} \tau' : \kappa$ and $\tEvalX{\tau'}{\tau'}$. 
-\end{theorem}
+%\begin{theorem}[Kind Safety and Termination]
+%If $\vdash_{\Sigma} \tau : \kappa$ then $\tEvalX{\tau}{\tau'}$ such that $\vdash_{\Sigma_0} \tau' : \kappa$ and $\tEvalX{\tau'}{\tau'}$. 
+%\end{theorem}
 
@@
 
-\begin{theorem}[Non-Interference]
-For any property $P(\gamma)$, if $\progOK{ }{\Sigma}{e}$ and $\ecompiles{ }{\fvalCtx}{e}{\tau}{\gabs}$ and $\eraseX{\gabs}{\gamma}$ implies $P(\gamma)$, then for any $\Sigma'$ and $\fvalCtx'$ disjoint from $\Sigma$ and $\fvalCtx$, we have that if $\progOK{ }{\Sigma,\Sigma'}{e'}$ and $\ecompiles{ }{\fvalCtx,\fvalCtx'}{e'}{\tau}{\gabs'}$ and $\eraseX{\gabs'}{\gamma'}$ then $P(\gamma')$.
-\end{theorem}
+%\begin{theorem}[Non-Interference]
+%For any property $P(\gamma)$, if $\progOK{ }{\Sigma}{e}$ and $\ecompiles{ }{\fvalCtx}{e}{\tau}{\gabs}$ and $\eraseX{\gabs}{\gamma}$ implies $P(\gamma)$, then for any $\Sigma'$ and $\fvalCtx'$ disjoint from $\Sigma$ and $\fvalCtx$, we have that if $\progOK{ }{\Sigma,\Sigma'}{e'}$ and $\ecompiles{ }{\fvalCtx,\fvalCtx'}{e'}{\tau}{\gabs'}$ and $\eraseX{\gabs'}{\gamma'}$ then $P(\gamma')$.
+%\end{theorem}
 The proof relies on the fact that the show/hide rules are mutually exclusive and so there are two cases to consider for each form of expression, and weakening properties of the family contexts.
@@
 \section{Use Cases}\label{examples}
+
+Aldrich argues that dynamic dispatch has proven useful for building  large software frameworks \cite{aldrich-onward13}.
+Object-oriented languages support dynamic dispatch directly but encodings of dynamic dispatch in terms of standard functional primitives can be awkward and difficult to reason about (e.g.
+\cite{pfpl} Ch.
+X).
+
 Indexed type families and indexed operator families are a powerful tool for programming language semantics.

### 2014-02-18 thesis-proposal/att (Stheno's conflicted copy 2014-02-17).tex
Further updates

% !TEX root = omar-thesis-proposal.tex
\input{../esop14/macros-atlam}
\input{macros-T}
\lstset{language=ML,
basicstyle=\ttfamily\footnotesize,
morekeywords={newcon,extends,tycon,opcon,err,schema,fix,is},
}

\section{Active Type Synthesis}\label{att}
General-purpose abstraction mechanisms, like those available in Wyvern and other languages today, are powerful, and should be used when possible. But relying on a small fixed collection of them may not always be satisfactory, for several reasons:
\begin{enumerate}
\item General-purpose abstraction mechanisms themselves are considerably varied. Tyconsider, for example, the many different ways languages expose product types: simple products, $n$-ary products, labeled products\footnote{We use the phrase \emph{labeled product} to avoid the connotations of related terms like \emph{record}, \emph{struct} and \emph{object}.}, structurally-typed labeled products, labeled products with mutable fields, labeled products with methods, labeled products with prototypic inheritance, labeled products with delegation, labeled products with class-based inheritance, and various combinations of these. New mechanisms and variants are constantly being developed. To allow these new general-purpose mechanisms to be evaluated in a controlled manner in practice, it must be possible to import them into existing codebases in a piecemeal fashion.
\item Specialized type systems that directly enforce stronger invariants than general-purpose abstraction mechanisms are capable of straightforwardly enforcing are often developed. In addition to the examples of type systems for regular expressions in Sec. \ref{regex} (\cite{a,b}), there are also specialized type systems for parallel programming \cite{a,b}, concurrent programming \cite{cml,a,b}, distributed programming \cite{tom7}, reactive programming \cite{reactiveml}, authenticated data structures \cite{popl13}, security policies \cite{walker00}, information flow \cite{smith2001}, web programming \cite{sandholm00}, aliasing \cite{naden12}, network protocols \cite{sekar99}, units of measure \cite{keneddy} and many others. These type systems are implemented by language-external means (e.g. by creating a new variant of ML, Java or C), leading to the problems of Sec. \ref{external-approaches}. 
\item Interoperability layers with external and legacy systems are another major area where support for directly implementing a new type system (that of the external language) and implementing its primitives in terms of a low-level interoperability API would be useful (e.g. \cite{CinJava}).
%\item General-purpose abstraction mechanisms are typically implemented uniformly. For example, inductive datatypes are typically implemented using tagged pointers to a payload. In some cases (e.g. natural numbers) it may be desirable to implement an abstraction in an alternative manner (e.g. as an integer) so that certain operations perform better. Parallel abstractions are of particular note, as they can benefit substantially from specialized implementation strategies on specialized hardware (e.g. a GPU).  Note that if control over internal implementation details were the only relevant use case, abstract datatypes might be appropriate.
\end{enumerate}

In this section, we will focus on language-integrated, type-oriented mechanisms for implementing extensions to the static and dynamic semantics of programming languages. We will leave the syntax fixed, inverting the setup in the previous section. We will begin by taking a  first-principles type-theoretic approach in Sec. \ref{foundations}, focusing on safety issues, and continue on in Sec. \ref{ace} by designing a full language called Ace. We will show how a range of statically-typed general-purpose abstraction mechanisms as well as more specialized abstractions can be expressed as safely-composable extensions within Ace. Notably, Ace is itself implemented as a library for Python. We discuss how to bootstrap a highly-expressive extensible static type system atop the quotation and object-oriented features available in Python, a uni-typed language.

\subsection{Foundations}\label{foundations}
Programming languages are typically organized around a monolithic collection of indexed type and operator constructors. Let us begin by considering the simply-typed lambda calculus (STLC). It provides a single type constructor, indexed by a pair of types, that we will call $\fvar{Arrow}$ for uniformity. It also provides two operator constructors: $\lambda$, indexed by a type, and $\texttt{ap}$, indexed trivially. We can specify the static semantics of the STLC as follows, using a uniform abstract syntax where type and operator indices are written in braces to emphasize this way of thinking about its structure:
\begin{figure}[h]
\small
\vspace{-8pt}
\begin{mathpar}
\inferrule[var]{ }{
	\jet{\eCtxX{\eCtx}{x}{\tau}}{x}{\tau}
}

\inferrule[arrow-I]{
	\jet{\eCtxX{\eCtx}{x}{\tau}}{e}{\tau'}	
}{
	\jetX{\lambda[\tau]({x}.e)}{\fvar{Arrow}[(\tau, \tau')]}
}

\inferrule[arrow-E]{
	\jetX{e_1}{\fvar{Arrow}[(\tau, \tau')]}\\
	\jetX{e_2}{\tau}
}{
	\jetX{\texttt{ap}[()](e_1; e_2)}{\tau'}
}
\end{mathpar}
\vspace{-8pt}
\caption{Static semantics of the simply typed lambda calculus.}
\end{figure}

Although a researcher may casually speak of ``conservatively extending the STLC`` with a primitive natural number type and its corresponding operators (as in G\"odel's T; Figure \ref{nat-statics}) or primitive product types and their  corresponding operators (Figure \ref{pair-statics}), it is clearly not possible to directly introduce these from within the STLC.
\begin{figure}[h]
\small
\vspace{-8pt}
\begin{mathpar}
\inferrule[nat-I1]{ }{
	\jetX{\texttt{z}[()]()}{\fvar{Nat}[()]}
}

\inferrule[nat-I2]{
	\jetX{e}{\fvar{Nat}[()]}
}{
	\jetX{\texttt{s}[()](e)}{\fvar{Nat}[()]}
}

\inferrule[nat-E]{
	\jetX{e_1}{\fvar{Nat}[()]}\\
	\jetX{e_2}{\tau}\\
	\jet{\eCtxX{\eCtxX{\eCtx}{x}{\fvar{Nat}[()]}}{y}{\tau}}{e_3}{\tau}
}{
	\jetX{\texttt{natrec}[()](e_1; e_2; x.y.e_3)}{\tau}
}
\end{mathpar}
\vspace{-8pt}
\caption{Static semantics of natural numbers.}\label{nat-statics}
\vspace{-15pt}
\end{figure}
\begin{figure}[h]
\small
\begin{mathpar}
\inferrule[Prod-I]{
	\jetX{e_1}{\tau_1}\\
	\jetX{e_2}{\tau_2}
}{
	\jetX{\texttt{pair}[()](e_1; e_2)}{\fvar{Prod}[(\tau_1, \tau_2)]}
}

\inferrule[Prod-E1]{
	\jetX{e}{\fvar{Prod}[(\tau_1, \tau_2)]}
}{
	\jetX{\texttt{prl}[()](e)}{\tau_1}
}

\inferrule[Prod-E2]{
	\jetX{e}{\fvar{Prod}[(\tau_1, \tau_2)]}
}{
	\jetX{\texttt{prr}[()](e)}{\tau_1}
}
\end{mathpar}
\vspace{-8pt}
\caption{Static semantics of products}\label{pair-statics}
\end{figure}

One recourse researchers have in such situations is to attempt to encode new constructs in terms of existing constructs. Collections of such encodings are sometimes called \emph{embedded domain-specific languages (EDSLs)} \cite{fowler2010domain}. Unfortunately, finding an encoding that fully captures both the static and dynamic semantics of a desirable construct can be difficult\footnote{For example, Church encodings in System F were among the more challenging topics for students in our undergraduate programming languages course, 15-312.} and is not always possible. Even when it possible, it may be impractical to work with and optimize. For our example of adding natural numbers or products to the STLC, a reasonable encoding is impossible. In closely-related languages, like the polymorphic lambda calculus (a.k.a. System F), such constructs are only weakly definable via Church encodings \cite{reynoldsIntroToPolymorphism} and far from practical. Reynolds, echoing Perlis   \cite{perlisPearl}, puts it succinctly \cite{reynoldsIntroToPolymorphism}:

\begin{quote}
To say that any reasonable function can be expressed by some program is not to say that it can be expressed by the most reasonable program. It is clear that the language requires a novel programming style.
Moreover, it is likely that certain important functions cannot be expressed by their most
efficient algorithms.
\end{quote}

%creating a new language. If this is not practical, the best one can attempt to do is encode the new types in terms of existing types (by a Church encoding, for example). This is generally unsatisfactory --

%Languages implemented using these common patterns are central planning by a language designer or design committee.

%Researchers or domain experts who cannot work around such limitations must develop new standalone languages. In our simple scenario, we may simply copy our implementation of G\"odel's T or even edit it directly (a pernicious technique for implementing a new language where the prior one is overwritten). In a more complex scenario, we may instead employ a tool like a compiler generator or DSL framework \cite{fowler2010domain} that can generate a standalone implementation from declarative specifications of language constructs. Some of these tools allow you to package and reuse these specifications (with the important caveat that not all combinations of constructs are valid and free of conflicts, an important modularity issue that we will return to several times in this paper).
%
%The increasing sophistication and ease-of-use of these tools have led many to suggest a {\it language-oriented approach} \cite{journals/stp/Ward94} to software development where different components of an application are written in different languages. Unfortunately, this leads to problems at language boundaries: a library's external interface must only use constructs that can reasonably be expressed in \emph{all possible calling languages}. This can restrict domain-specific languages by, for example, precluding constructs that rely on statically-checked invariants stronger than those their underlying representation in a common target language normally supports. At best, constructs like these can be exposed by generating a wrapper where run-time checks have been inserted to guarantee necessary invariants. This compromises both verifiability and performance and requires the development of an interoperability layer for every DSL. Moreover, library clients must work with verbose and unnatural ``glue code'' when interfacing across languages, defeating the primary purpose of high-level programming languages: hiding the low-level details from the end-users of abstractions. We diagram this fundamental \emph{compatibility problem} in Figure \ref{approaches}(a).
%\begin{figure*}
%\begin{center}
%\includegraphics[scale=0.5]{approaches.pdf}
%\end{center}
%\vspace{-20px}
%\caption{\small (a) With a language-oriented approach, novel constructs are packaged into separate languages. Users can only safely and naturally call into languages consisting of common constructs (often only the common target language, such as C or Java bytecode). (b) With a language-internal extensibility approach, there is one system providing a common internal language, where additional primitive constructs that strictly strengthen its static guarantees or perform specialized code generation are specified and distributed within libraries. \label{approaches}}
%\end{figure*}

%As a result, domain-specific languages and new general-purpose abstractions alike have experienced relatively slow adoption in practice.
%
%Porting large codebases to new languages is difficult, and the dominant programming languages innovate slowly, so programming language.
%
%More specifically, such languages are neither \emph{internally extensible} because the language itself exposes only natural numbers and functions to its users, nor are they \emph{externally extensible} because no new behaviors can be added to the language's  implementation in a separate module from the one containing the initial implementation.

%This is the essence of a monolithic language implementation: it is impossible for anyone to modularly extend languages defined in this way.

An extensible programming language could address these problems by providing a language-integrated mechanism for introducing new type and operator constructors and implementing their associated static and dynamic semantics directly. 
%Library developers need only consider which abstractions are most appropriate for their domain, without also considering whether these constructs can be exposed using abstractions appropriate to the domains of client code. Clients can simply import any necessary constructs when using a library that relies on them, preserving safety and ease-of-use without the use of  wrappers and glue code. We show this competing approach in Figure \ref{approaches}(b).
%Researchers and domain experts thus gain the ability to distribute new ideas for evaluation to a broader development community without requiring the approval of maintainers of mainstream languages, large-scale porting of code or explicit interoperability layers. 
But, as mentioned in Section \ref{language-integrated-approaches}, some significant challenges must be addressed before such a mechanism can be relied upon. The desire for expressiveness must be balanced against  concerns about maintaining various safety properties in the presence of arbitrary combinations of user-defined  extensions to the language's core semantics. The mechanism must ensure that desirable \emph{metatheoretic properties} (e.g. type safety, decidability) of the language are maintained by extensions. Because multiple independently developed extensions might be used within one program, the mechanism must further guarantee their \emph{non-interference}. These are the issues we seek to address in this work.
%\subsection{Theory}\label{atlam}
%%\begin{figure}
%%\begin{mathpar}
%%\inferrule{a}{b}
%%\end{mathpar}
%%\end{figure}
%In this section, we will describe a minimal calculus that captures our language-internal extensibility mechanism,  called \emph{active typechecking and translation (AT\&T)}. AT\&T allows developers to declare new primitive type families, associate operators with them, implement their static semantics in a functional style, and realize their dynamic semantics by simultaneously implementing a translation into a typed internal language. Note that this latter mechanism is closely related to how Standard ML was (re-)specified\todo{cite/read a bit about this/ask Bob/flesh this out}, but that we are fundamentally interested in extending language \emph{implementations}, not their  declarative specifications; proving the adequacy of such implementations against mechanized specifications, or extracting them directly from such specifications, will be investigated in future work.
%
%The AT\&T mechanism utilizes type-level computation of higher kind and integrates typed compilation techniques into the language to allow us to give strong metatheoretic guarantees,  and uses a mechanism notionally related to abstract types (such as those found in the ML module system) to guarantee that extensions cannot interfere with one another,  while remaining straightforward and expressive. In this section, we will develop a core calculus, called \atlam, which uses a minimal, uniform grammar for primitive operators. Then in Section \ref{ace}, we will show how to realize this minimal mechanism within a widely-used language with a more expressive grammar.
%
%AT\&T is general with respect to many choices about the type-level language, the typed internal language and syntax. Choices along these dimensions can affect both expressiveness and ease-of-use. We will begin in Sec. 2 by introducing a minimal system called $@\lambda$ (the ``actively-typed lambda calculus'') that distills the essence of the mechanism in a simply-typed, simply-kinded setting. This will allow us to fully and concisely formalize the language and compiler and give several key safety theorems. We will then continue in Sec. 3 by discussing variants of this mechanism based on other basic paradigms, considering dependently-typed functional languages and object-oriented languages, discussing trade-offs between expressivity and safety when doing so. We have developed a simple prototype called Ace and have used it to develop a number of full-scale language extensions as libraries. We will briefly discuss this language and these extensions in Sec. 4.

%We note at the outset that AT\&T focuses on extending the static semantics of languages with fixed, though flexible, syntax. Language-internal syntax extension mechanisms have been developed in the past (e.g. SugarJ \cite{sugarj}) but they have also suffered from safety problems because grammar composition is not always safe when done in an  unconstrained manner. Tyconstrained approaches that provide stronger safety guarantees have recently been outlined (e.g. Wyvern \cite{globaldsl13}) but we will leave integration of syntax extensions with semantic extensions as future work.

\subsubsection{From Extensible Compilers to Extensible Languages}\label{evolution}
The monolithic character of most programming languages is reflected in, and perhaps influenced by, the most common mechanisms used for implementing programming languages.
Let us consider an implementation of the STLC. 
A compiler written using a functional language will invariably represent the primitive type  and operator constructors using {closed} recursive sums. 
A simple implementation in Standard ML could be based around these datatypes, for example:
\begin{lstlisting}
  datatype Type = Arrow of Type * Type
  datatype Exp = Var of var | Lam of var * Type * Exp | Ap of Exp * Exp 
\end{lstlisting}

The compiler front-end, which consists of a typechecker and translator to a suitable intermediate language (for subsequent compilation by some suitable back-end), will proceed by exhaustive case analysis over the constructors of \lstinline{Exp}.

In a class-based object-oriented implementation of Godel's T, we might instead encode type and operator constructors as subclasses of abstract classes \lstinline{Type} and \lstinline{Exp}. Typechecking and translation, however, will proceed by the ubiquitous \emph{visitor pattern}, dispatching against a fixed collection of {known} subclasses of \lstinline{Exp}.

In either case, we encounter the same basic issue: there is no way to modularly add new primitive type and operator constructors to \li{Type} and \li{Exp}, respectively, and provide implementations of their associated typechecking and translation logic. Everything is defined in precisely one place.
%This issue is related to the widely-discussed \emph{expression problem} (in a restricted sense -- we do not consider adding new functions beyond typechecking and translation here, only adding logic to these) \cite{wadler-expression}.

A number of language mechanisms have been proposed that allow new cases to be added to datatypes and the functions that operate over them in a modular manner. 
In functional languages, we might use \emph{open datatypes} \cite{opendatatypes}\todo{cite}. For example, if we wish to extend our implementation of the STLC with product types and we have written our compiler in a language supporting open datatypes, it might be possible to add new constructors: 
\begin{lstlisting}
  newcon Prod of Type * Type extends Type
  newcon Pair of Exp * Exp extends Exp
  newcon PrL of Exp extends Exp
  newcon PrR of Exp extends Exp
\end{lstlisting}

The logic for functionality like typechecking and translation can then be implemented for only these new cases. For example, the \lstinline{typeof} function that synthesizes a type for an expression given a context could be extended to support the new case \li{PrL} like so:
\begin{lstlisting}
  typeof (ctx, PrL(e)) = case typeof (ctx, e) of 
      Prod(t1, _) => t1 
    | _ => raise TypeError("<appropriate error message>")
\end{lstlisting}

If we allowed users to define new modules containing definitions like these and link them into our compiler, we will have succeeded in creating an externally-extensible compiler, albeit one where safety and non-interference is not guaranteed (we will return to this point shortly). We have not, however, created an extensible programming language because other compilers for the same language will not necessarily support the same extensions. 
If our newly-introduced constructs are exposed at a library's  interface boundary, clients of that library who are using different compilers face the same problems with client compatibility that those using different languages face (as described in Sec. \ref{external-approaches}). That is, {extending a language by extending a single compiler for it is morally equivalent to creating a new language}.

We argue that a more appropriate and useful place for extensions like this is directly within libraries, alongside abstractions that can be implemented in terms of existing primitive abstractions. To enable this, the language must allow for the introduction of new type constructors, like \lstinline{Prod}, operator constructors, like \lstinline{Pair}, \lstinline{PrL} and \lstinline{PrR}, and their associated type synthesis and translation logic. Because this mechanism is integrated into the language specification, all compilers must support it.

The design described above suggests we may now need to add another layer to our language, above the type-level language (conventionally, $\tau$) and expression language (conventionally, $e$), where extensions are implemented. In fact, we will show that \textbf{a natural place for type system extensions is within the type-level language}. The intuition  is that extensions to a statically typed language's semantics will need to manipulate types as values at compile-time. Many languages already allow users to write type-level functions for various reasons, effectively supporting this notion of types as values at compile-time. The type-level language is often constrained by its own type system, where the types of type-level terms are called \emph{kinds} for clarity,  that prevents type-level functions from causing problems during compilation. The kind system is often more constraining than the type system is (e.g. it might permit only terminating functions, to ensure that typechecking is decidable). This is precisely the structure that a distinct extension layer would have, and so we will show that is quite natural to unify the two.

\subsubsection{@$\lambda$: Conservatively Extending a Type System From Within}
\begin{figure}[t]
\small
$$\begin{array}{rccl}	
\textbf{programs} & \rho & ::= & \pfam{\familyDf}{\progsort} \pipe \pdef{t}{\kappa}{\tau}{\progsort} \pipe e\\
\text{operators}		&	\theta	&	::= &	\tops{op}{\kappaidx}{i}{a}{\taut{def}} \pipe 
												\topp{\theta}{\theta}\\
\\
\textbf{external terms} 				&	e	&	::=	&	\evar{x} \pipe 
														\elam{\evar{x}}{\tau}{e} \pipe 
														\eop{Tycon}{op}{
															\tauidx
														}{
  												    		\splat{e}{1}{n}
														} \\
									& 		&		& 	\\

\hspace{-5pt}\textbf{type-level terms} 	& \tau 	& ::= 	& 	\tvar{t} \pipe 
														\tlam{t}{\kappa}{\tau} \pipe 
														\tapp{\tau_1}{\tau_2} \pipe
														\tnil{\kappa} \pipe \tcons{\tau_1}{\tau_2} \pipe 
									                     \tfold{\tau_1}{\tau_2}{h}{t}{r}{\tau_3}
														\\

\text{type-level data}	 			& 		& \pipe	& 	 \iintlit \pipe \iop{\tau_1}{\tau_2} \pipe \tstr{str} \pipe \tunit \pipe 
														\tpair{\tau_{1}}{\tau_{2}} \pipe 
														\tfst{\tau} \pipe 
														\tsnd{\tau} 
														\\

\text{types} 						& 		& \pipe	& 	\ttypestd \pipe \tfamcase{\tau}{Tycon}{x}{\tau_1}{\tau_2}\\
						%				& & \pipe & \tfamcase{\tau}{Fam}{x}{\tau_1}{\tau_2}\\
\text{structural equality}  & & \pipe & 					\tifeq{\tau_{1}}{\tau_{2}}{\kappa}{\tau_{3}}{\tau_{4}} 
														\\																																					
\text{elaborations} 				& 		 & 	\pipe	&	\tden{\tauiterm}{\tautype} \pipe \terr \pipe \tdencase{\tau}{x}{t}{\tau_1}{\tau_2}\\
 %& & \pipe & 
%														\tdencase{\tau}{y}{x}{\tau_1}{\tau_2}
%														 \\

\text{reified IL}		&		&	\pipe	&	\titerm{\iota} \pipe \titype{\sigma} \\

												\\
\textbf{kinds} 					& \kappa	&	::=	&	\karrow{\kappa_1}{\kappa_2} \pipe \klist{\kappa} \pipe \dint \pipe
											    \kstr \pipe
												\kunit \pipe 
												\kpair{\kappa_{1}}{\kappa_{2}} \pipe 
												\kTypeBlur \pipe \kDen \pipe 
												\kITerm \pipe \kIType
												\\
\\												
%\textbf{ops signature}			& \Theta	&	::=	&	\kOpEmpty \pipe \kOp{\Theta}{op}{\kappai}\\
%											 							&		&		&	\\
\textbf{internal terms} 				& 	\iota	&	::=	&	\evar{x} \pipe 
												\ilam{\evar{x}}{\sigma}{\iota} \pipe 
												\iapp{\iota_{1}}{\iota_{2}} \pipe
												\ifix{\evar{f}}{\sigma}{\iota} \pipe
												\ipair{\iota_{1}}{\iota_{2}} \pipe 
												\ifst{\iota} \pipe
												\isnd{\iota}
												\\
							& 		& 	\pipe	& 
												\iintlit \pipe \iop{\iota_{1}}{\iota_{2}} \pipe \iIfEq{\iota_{1}}{\iota_{2}}{\dint}{\iota_{3}}{\iota_{4}}  
												\\
 &  & \pipe & \tvalof{\tau_1}{\tau_2} \pipe \iup{\tau} \\
%\text{deabstracted}& \iota & ::= & \mathcal{G}[\iota, \sigma]\\
\textbf{internal types}			&	\sigma	&	::=	&    \darrow{\sigma_1}{\sigma_2} \pipe
												\dint \pipe
												\dpair{\sigma_1}{\sigma_2} \pipe
												\trepof{\tau} \pipe \dup{\tau}\\
\end{array}$$
\vspace{-10pt}
\caption{\small Abstract syntax of \atlam. Here, $x$ ranges over external and internal language variables, $\tvar{t}$ ranges over type-level variables, $\fvar{Tycon}$ ranges over type constructor names and $\opvar{op}$ ranges over operator constructor names. The form $\tstr{str}$ represents string literals, $\iintlit$ represents integer literals and $\oplus$ represents standard total binary operations over integers.%The productions related to the internal language are written using generators $\mathcal{G}$ and $\mathcal{S}$ to avoid duplicating the syntax of common terms.
\label{grammar}}
\end{figure}
\begin{figure}[t]
\begin{lstlisting}
tycon Nat of 1 with  
    schema $\lambda$idx:1.$\blacktriangledown$($\mathbb{Z}$) 
    opcon Z of 1 ($\lambda$idx:1.$\lambda$args:list[Elab].is_empty args $\llbracket$$\triangledown$(0) as Nat[()]$\rrbracket$)
    opcon S of 1 ($\lambda$idx:1.$\lambda$args:list[Elab].pop_final args $\lambda$x:ITm.$\lambda$ty:$\star$. 
      checktype ty Nat[()] $\llbracket$$\triangledown$($\vartriangle$(x)+1) as Nat[()]$\rrbracket$)
    opcon Rec of 1 ($\lambda$idx:1.$\lambda$args:list[Elab].
      pop args $\lambda$x1:ITm.$\lambda$ty1:$\star$.$\lambda$args':list[Elab].
      pop args' $\lambda$x2:ITm.$\lambda$ty2:$\star$.$\lambda$args'':list[Elab].
      pop_final args'' $\lambda$x3:ITm.$\lambda$ty3:$\star$.
      check_type t1 Nat[()] (
      check_type t3 Arrow[(Nat[()], Arrow[(t2, t2)])] 
      	$\llbracket$$\triangledown$((fix f:$\mathbb{Z} \rightarrow$ rep(t2) is $\lambda$x:$\mathbb{Z}$.
	        if x = 0 then $\vartriangle$(x2) else $\vartriangle$(x3) (x - 1) (f (x - 1))) $\vartriangle$(x1)) as t2$\rrbracket$))
end
\end{lstlisting}
\caption{An implementation of primitive natural numbers as internal integers in @$\lambda$.}
\label{nat-atlam}
\end{figure}
We will begin by developing an ``actively-typed'' version of the simply-typed lambda calculus with type-level computation called @$\lambda$. In the semantics for this calculus, we will integrate typed compilation techniques and a form of type abstraction to allow us to prove key type safety and non-interference theorems. The abstract syntax of @$\lambda$ is in Fig. \ref{grammar}.

The semantics of the language are in the style of the Harper-Stone elaboration semantics for Standard ML: all terms in an external language, $e$, are simultaneously assigned a type and an elaboration to a term, $\iota$, in a fixed typed internal language \cite{HarperStone, Stone}. The mechanism checks that any generated elaborations are \emph{type preserving} -- that all external terms of a particular type, $\tau$, elaborate to internal terms with a consistent internal type, $\sigma$ -- following the approach taken in the TIL compiler for Standard ML by Morrisett et al. \cite{TIL}. This allows the semantics to be reasoned about compositionally and, when combined with type safety of the internal language, gives us type safety of the extensible external language. 
This design ensures that global properties that the internal language maintains are guaranteed to hold no matter which extensions are introduced. In other words, extension providers cannot implement type systems weaker than the type system of the internal language. Providers are, however, able to implement type systems that maintain invariants stronger than those the internal type system can maintain. The internal language of @$\lambda$ is PCF with simple products and one base type, integers. Because of the availability of the fixpoint operator, the internal language is universal (i.e. Turing-complete).

It is reasonable, if we wish to actually implement this language, to permit only the addition of deterministic, decidable rules to the semantics. Extension logic is thus implemented in a functional style  (essentially, lifting fragments of the typechecker into the type-level language), rather than extracted from inductive specifications, like those in the Sec. \ref{foundations}\footnote{Though we will not explore this further in this thesis, a system for extracting such a function from a declarative specification could potentially be implemented using the techniques in Sec. \ref{aparsing} -- a \emph{kind-specific language}!}. This has the added benefit of giving providers finer-grained control over error reporting. When extracting an implementation from a typical inductive specification, which only specifies ``success conditions'', it is difficult to provide different error messages at different failure points within a single rule.
%In this section, we will sketch a calculus where new type families and operator families can be introduced by users. The type synthesis and translation logic that implements the desired static and dynamic semantics, respectively, for these is implemented in the type-level language.

To give a first example, the code in Fig. \ref{nat-atlam} (written using concrete syntax) shows how to introduce a new type constructor, $\fvar{Nat}$, indexed trivially (i.e. by a type-level value of kind 1, also known as unit). Types themselves are type-level values of kind $\kTypeBlur$ and are introduced by naming a type constructor and providing an index of the appropriate kind: $\ttype{Nat}{\tunit}$. To ensure that type equality is decidable, type constructors can only be indexed by values of a kind for which equality is ``trivially'' decidable (i.e. where semantic equality coincides with syntactic equality). This means that types themselves, as well as type-level data structures containing values of equality kind, can be used, while type-level functions cannot.

To ensure that compilation is type-preserving, as described above, each type must have an internal type, called its \emph{representation}, associated with it. The representation of a type might depend on its index, so each type constructor must declare a type-level function that returns an internal type given an index, called its \emph{representation schema}. An internal type, $\sigma$, is reflected into a type-level terms of kind $\kIType$ via the introductory form $\blacktriangledown(\sigma)$. Because $\fvar{Nat}$ is indexed trivially, it simply returns the reflected form of the internal type $\mathbb{Z}$. %The sort $\sigma$ also contains forms $\trepof{\tau}$ and $\dup{\tau}$. The former allows one to refer, abstractly, to the representation of the type $\tau$, while the latter enables the interpolatation of a type-level value of kind $\kIType$.

%Two additional forms, $\tvalof{\tauiterm}{\tautype}$ and $\iup{\tau}$, will be discussed below. These are erased by the end of the elaboration process.

In the external language, $e$, all operators (other than $\lambda$, for reasons we will discuss) are invoked uniformly. The natural number two, for example, is written: $$\eop{Nat}{s}{()}{\eop{Nat}{s}{()}{\eop{Nat}{z}{()}{ }}}$$

Operator constructors, like type constructors, are indexed by type-level values. In our example, all the operators are indexed trivially, but we will see examples later where non-trivial operator indices are useful. We will discuss why operator constructors are associated directly with type constructors, rather than existing as standalone entities, when we discuss the scoping mechanism for the representation schema, which is essential for guaranteeing non-interference. This coupling also forms the foundation for a higher-level elaboration mechanism that allows us to use more conventional and concise (though still fixed) syntactic idioms, which we intend to explore in this thesis, but have not yet formalized. For now, we simply note that intuitively, most operators are either introduction or elimination forms for a particular type, so we are simply encoding a common idiom.

The declarations of the three operator constructors related to natural numbers are given on lines 3-13 of Fig. \ref{nat-atlam}. Operator constructors are declared by specifying the kind of type-level value they are indexed by and giving an \emph{operator definition}, a type-level function of kind $\kappa \rightarrow \klist{\kDen} \rightarrow \kDen$. The kind $\kDen$ classifies \emph{elaborations}. There are two introductory forms for this kind: \emph{valid elaborations}, $\tden{\tauiterm}{\tautype}$, consist of a reflected internal language term paired with a type assignment, while \emph{error elaborations}, $\terr$, represent a type error (in a full implementation, the error elaboration would contain an error message, as discussed above, but for simplicity in our core calculus, we simply treat all type errors equivalently). Internal terms, $\iota$, like internal types, are reflected as type-level terms of kind $\kITerm$ via the introductory form $\titerm{\iota}$. The form $\iup{\tau}$ interpolates a type-level value of kind $\kITerm$ into an internal term.  Note that there are no elimination forms for the kinds $\kIType$ and $\kITerm$. Composition occurs entirely via interpolation -- syntax trees are never examined directly by extensions in @$\lambda$.

Operator definitions are called by the compiler to generate an elaboration for operator invocations. In our example, the definition of $\fvar{Nat}.\opvar{z}$ simply checks that there were no arguments (using a simple helper function for working with lists of elaborations, \li{is_empty}, not shown) and if so, provides an elaboration where the internal language integer \li{0} is associated with type $\ttype{Nat}{\tunit}$. If the list is not empty, \li{is_empty} simply returns $\terr$. The other two operator constructors are more verbose, but follow a similar pattern -- elaborations are popped off and broken down into their constituent internal terms and types by helper functions, then a functional program that implements the intended semantics is written to produce an elaboration as a whole. Note that the \li{Rec} operator does not itself bind new variables; it must use the built-in \li{Arrow} type constructor. There is no support in @$\lambda$ for introducing new forms of binding; $\lambda$ is the only available binding operator. Note that application is invoked like other operators: $\eop{Arrow}{ap}{\tunit}{e_1; e_2}$.

This definition of natural numbers can be shown to be an adequate encoding of the semantics given in Sec. \ref{foundations}. It can also be shown to maintain the invariant that natural numbers elaborate into \emph{non-negative} internal integers. Note that because we are beginning with a simply-typed, simply-kinded formulation, these kinds of statements must be proven metatheoretically (as with other sorts of complex invariants in simply-typed languages). With a naive formulation of this mechanism, these theorems would be quite a bit more precarious, however. If another provider, perhaps a malicious one, declares a new operator constructor that introduces natural numbers, these theorems may not be \emph{conserved}. For example, the following operator declaration is quite problematic:

\begin{lstlisting}
opcon badnat of 1 ($\lambda$idx:1.$\lambda$args:list[Elab].$\llbracket\triangledown$(-1) as Nat[()]$\rrbracket$)
\end{lstlisting}

With this declaration, there is now a new and unexpected introductory form for the natural number type, and even worse, it violates our previous implementation invariants. A naive check for type preservation would not catch this problem: \li{-1} does have type $\mathbb{Z}$, it is simply not an integer that could have been emitted by the original definition of \li{Nat}.

To prevent this problem, we \emph{hold the representation of a type abstract} outside of the operators explicitly associated with it. If \li{badnat} cannot know that the representation of \li{Nat[()]} is $\mathbb{Z}$, then the above operator implementation cannot be correct. Because we cannot prove relational correctness properties from within a simply-typed/kinded calculus, our semantics enforces this by using a form of value abstraction, similar to that described by Zdancewic et. al \cite{zdancewic}. In brief: the representation of a type, written $\trepof{\tautype}$, does not reduce further to the actual representation type (e.g. to $\mathbb{Z}$) when not in an operator definition associated with $\tautype$. The only internal form that can be checked against  $\trepof{\tautype}$ is the special form $\tvalof{\tauiterm}{\tautype}$, an abstracted internal term corresponding to the internal term reflected by $\tauiterm$. Rather than exposing it explicitly, it is wrapped so that it can't be checked against any type other than $\trepof{\tautype}$. A full description of this mechanism (and the others, described above) requires us to introduce the semantics in detail. A draft of the semantics of @$\lambda$ is available as a paper draft\footnote{\url{https://github.com/cyrus-/papers/tree/master/esop14}}.

The use cases permitted by this mechanism are not simply subsumed by a module system with support for abstract types and functors. Indeed, such a module system is an orthogonal concern in that it must sit atop a core type system. More particularly, the distinction is the fundamental one between functions and operators. Functions cannot examine type indices at compile-time to determine a return type and implementation (or generate static error messages), as operators are able to do. It is encouraging that there are clear parallels between module systems with abstract types and type constructors with abstract schemas, and any full-scale language design based on this mechanism should certainly provide both mechanisms (with the former used in most cases). We plan to investigate this relationship more thoroughly, showing examples where abstract types are more clearly unsuitable, in the remainder of this work.

\subsubsection{Remaining Tasks and Timeline}
We are actively working on this mechanism and plan to submit a paper to ICFP 2014 on Mar. 1 based on a submission to ESOP 2014 that was not accepted. The following key tasks, other than clarifying the writing, remain to be completed:

\begin{enumerate}
\item Reviewers of the previous submission to ESOP found a problem with our treatment of internal type variables that needs to be corrected by threading a context through the type-level evaluation semantics.
\item We must more rigorously state and prove the key lemmas and theorems related to type preservation, type safety, decidability and conservativity. 
\item We must develop examples that are more interesting than natural numbers. In particular, some interesting form of labeled product type that SML doesn't have (e.g. a record with prototypic dispatch) as well as sum types would be interesting.
\item We must consider whether properties related to termination can be conserved by this mechanism, because a fixpoint can be used to introduce a non-terminating expression of any type. This may be possible by being careful about how deabstraction interacts with fixpoints.
\item We must more clearly connect this work to the Harper-Stone semantics and other related work, and clarify  the relationship to abstract types.
\item We must show how to add additional syntactic forms to the external language that defer to the generic operator invocation form in a type-directed manner. This connects the theory to the work on Ace, described below. 
\end{enumerate}

\subsection{Ace}\label{ace}
\lstset{
  language=Python,
  showstringspaces=false,
  formfeed=\newpage,
  tabsize=4,
  commentstyle=\itshape,
  basicstyle=\ttfamily\scriptsize,
  morekeywords={lambda, self, assert, as},
  numbers=left,
  numberstyle=\scriptsize\color{light-gray}\textsf,
  xleftmargin=2em,
  stringstyle=\color{mauve}
}
\lstdefinestyle{Bash}{
    language={}, 
    numbers=left,
    numberstyle=\scriptsize\color{light-gray}\textsf,
    moredelim=**[is][\color{blue}\bf\ttfamily]{`}{`}
}
\lstdefinestyle{OpenCL}{
	language=C++,
	morekeywords={kernel, __kernel, global, __global, size_t, get_global_id, sin, printf, int2}
}

%\usepackage{float}
%\floatstyle{ruled}
%\newfloat{codelisting}{tp}{lop}
%\floatname{codelisting}{Listing}
%\setlength{\floatsep}{10pt}
%\setlength{\textfloatsep}{10pt}

The key choices that a language must make when supporting the mechanism described in the previous section are:
\begin{enumerate}
\item What is the semantics of the type-level language?
\item What is the syntax of the external language, and how should the syntactic forms dispatch to  operator definitions?
\item What is the semantics of the internal language?
\end{enumerate}

In @$\lambda$, the type-level language was a form of the simply-typed lambda calculus with a few common data types, and the internal language was a variant of PCF. The external language currently only includes direct dispatch by naming an operator explicitly. This is useful for the purposes of clarifying the foundations of our work.

In this section, we wish to explore a different point in this design space, with an eye toward practicality and expressiveness. In particular, we want to show how to implement {the extension mechanism itself} as a library within an existing, widely used language, solving a \emph{bootstrapping problem} that has prevented  other work on extensibility from being effective as a means to bring more research into practice. This language, called Ace, makes the following choices:

\begin{enumerate}
\item Python is used as the type-level language (and more generally, as a compile-time metalanguage).
\item Python's syntax is used for the external language. We will describe the dispatch protocol below.
\item The internal language can be user-defined, rather than being pre-defined as in @$\lambda$.
\end{enumerate}

The choice of Python as the host language presents several challenges because we are attempting to embed an extensible static type system within a uni-typed (a.k.a. dynamically typed) language, without modifying its syntax in any way. We show that by leveraging Python's support for function quotations and by using its class-based object system to encode type constructors, we are able to accomplish this goal. Then, with a flexible statically-typed language under the control of libraries, we implement a variety of statically-typed abstractions, including common functional abstractions (e.g. inductive datatypes with pattern matching), low-level parallel abstractions (all of the OpenCL programming language), object systems and domain-specific abstractions (e.g. regular expression types, as described in the introduction). Ace can be used both as a standalone language and as a staged compilation environment from within Python.

\subsubsection{Language Design and Usage}\label{usage}
Listing \ref{map} shows an example of an Ace file. As promised, the top level of an Ace file is written directly in Python, requiring no modifications to the language (versions 2.6+ or 3.3+) nor features specific to CPython (so Ace supports alternative implementations like Jython, IronPython and PyPy). This choice pays  dividends on line 1: Ace's package system is Python's package system, so Python's build tools (e.g. \verb|pip|) and package repostories (e.g. \verb|PyPI|) are directly available for distributing Ace libraries.

The top-level statements in an Ace file, like the \verb|print| statement on line 10, are executed at compile-time. That is, Python serves as the \emph{compile-time metalanguage} of Ace. %For readers familiar with C/C++, Python can be thought of as serving a role similar to (but more general than) its preprocessor and template system (as we will see).
Functions containing run-time behavior, like \verb|map|, are annotated as Ace functions and are then governed by a semantics that differs substantially from Python's (in ways that we will describe below). But Ace functions share Python's syntax. As a consequence, users of Ace benefit from an ecosystem of well-developed tools that work with Python syntax, including parsers, code highlighters, editor modes, style checkers and documentation generators.

\subsubsection{OpenCL as an Active Library}
The code in this section uses \verb|clx|, an example of an library that implements the semantics of the OpenCL programming language, and extends it with some additional useful types, using Ace. Ace itself has no built-in support for OpenCL.

To briefly review, OpenCL provides a data-parallel SPMD programming model where developers define functions, called {\em kernels}, for execution across thousands of threads on \emph{compute devices} like GPUs or multi-core CPUs \cite{opencl11}. Each thread has access to a unique index, called its \emph{global ID}. Kernel code is written in the OpenCL kernel language, a somewhat simplified variant of C99 extended with some new primitive types and operators, which we will describe as needed in our examples below.

\subsubsection{Generic Functions}\label{genfn}
\begin{codelisting}
\lstinputlisting{../ace-pldi14/listing3.py}
\caption{[\texttt{listing\ref{map}.py}] A generic imperative data-parallel higher-order map function targeting OpenCL.}
\label{map}
\end{codelisting}
%\begin{codelisting}
%\lstinputlisting[commentstyle=\color{mauve}]{../ace-pldi14/listing7.py}
%\caption{[\texttt{listing\ref{metaprogramming}.py}] Metaprogramming with Ace, showing how to construct generic functions from abstract syntax trees.}
%\label{metaprogramming}
%\end{codelisting}
%\begin{codelisting}
%\begin{lstlisting}[style=Bash]
%$ `acec listing3.py`
%Hello, compile-time world!
%[ace] TypeError in listing1.py (line 6, col 28): 
%      'GenericFnType(negate)' does not support [].
%[acec] listing3.cl successfully generated.
%\end{lstlisting}
%\caption{Compiling \texttt{listing\ref{compscript}.py} using the \texttt{acec} compiler.}
%\label{mapc}
%\end{codelisting}
%\begin{codelisting}
%\lstinputlisting[style=OpenCL]{listing5.cl}
%\caption{[\texttt{listing\ref{compscript}.cl}] The OpenCL file generated by Listing \ref{mapc}.}
%\label{mapout}
%\end{codelisting}

Lines 3-4 introduce \verb|map|, an Ace function of three arguments that is governed by the \emph{active base} referred to by \verb|clx.base| and targets the \emph{active target} referred to by \verb|clx.opencl|. The active target determines which language the function will compile to (here, the OpenCL kernel language) and mediates code generation. The active target plays an analagous role to the internal language of @$\lambda$.

The body of this function, on lines 5-8, does not have Python's semantics. Instead, it will be governed by the active base together with any \emph{active types} used within it. No  types have yet been assigned, however. Because our type system is extensible, the code inside could be meaningful for many different assignments of types to the arguments (a form of \emph{ad hoc polymorphism}). We call functions awaiting types, like \verb|map|,  \emph{generic functions}. Once types have been assigned, they are called \emph{concrete functions}.

Generic functions are represented at compile-time as instances of \verb|ace.GenericFn| and consist of an abstract syntax tree, an {active base}, an {active target} and a read-only copy of the Python environment that they were defined within. The purpose of the \emph{decorator} on line 3 is to replace the Python function on lines 4-8 with an Ace generic function having the same syntax tree and environment and the provided active base and active target. 
A decorator in Python is simply syntactic sugar that applies another function directly to the function  being decorated \cite{python}. In other words, line 3 could be replaced by the following  statement on line 9: \verb|map = ace.fn(clx.base, clx.opencl)(map)|.
Ace extracts the abstract syntax tree for \verb|map| using the Python standard  library packages  \verb|inspect| (to retrieve its source code) and \verb|ast| (to parse it into a syntax tree). The ability to extract a function's syntax tree and inspect its closure directly are the two key ingredients for implementing a mechanism like this as a library within another language.

\subsubsection{Concrete Functions and Explicit Compilation}
\begin{codelisting}
\begin{lstlisting}
import listing1, ace, examples.clx as clx

@ace.fn(clx.base, clx.opencl)
def negate(x):
  return -x

T1 = clx.Ptr(clx.global_, clx.float)
T2 = clx.Ptr(clx.global_, clx.Cplx(clx.int))
TF = negate.ace_type

map_neg_f32 = listing1.map[[T1, T1, TF]]
map_neg_ci32 = listing1.map[[T2, T2, TF]]
\end{lstlisting}
%\lstinputlisting{../ace-pldi14/listing4.py}
\caption{[\texttt{listing2.py}] The generic \texttt{map} function compiled to map the \texttt{negate} function over two  types of input.}
\label{compscript}
\end{codelisting}

To compile a generic function to a particular \emph{concrete function}, a type must be provided for each argument, and typechecking and elaboration must then succeed. Listing \ref{compscript} shows how to explicitly provide type assignments to \verb|map| using the subscript operator (implemented using Python's operator overloading mechanism). We do so two times in Listing \ref{compscript}, on lines 11 and 12. Here, \verb|T1|, \verb|T2|, \verb|TF|, \verb|clx.float|, \verb|clx.int| and \verb|negate.ace_type| are types and \verb|clx.Ptr| and \verb|clx.Cplx| are type constructors. We will discuss these in the next section.

Concrete functions like \verb|map_neg_f32| and \verb|map_neg_ci32| are instances of \verb|ace.ConcreteFn|. They consist of a \emph{typed} abstract syntax tree, an elaboration into the target language and a reference to the originating generic function. %The typing and translation process is mediated by the logic in the active base, types and target that have been provided, as we will describe in more detail below.

To produce an output file from an Ace ``compilation script'' like \verb|listing|\texttt{\ref{compscript}}\verb|.py|, the command \verb|acec| can be invoked from the shell, as shown in Listing \ref{mapc}. The result of compilation is the OpenCL file shown in Listing \ref{mapout}. The \verb|acec| compiler (a simple Python script) operates in two stages:
\begin{enumerate}
\item Executes the provided Python file (\verb|listing3.py|).
\item Extracts the elaborations from concrete functions and other top-level constructs that define elaborations (e.g. types requiring declarations) in the final Python environment.  This may produce one or more files, depending on which active targets were used (here, just \verb|listing3.cl|, but a web framework built upon Ace might produce separate HTML, CSS and JavaScript files).
\end{enumerate}

%In this case, stage 1 results in the output on lines \ref{mapc}.2-\ref{mapc}.4. The type error printed on lines \ref{mapc}.3-\ref{mapc}.4 will be explained in the next section. The compiler then enters stage 2 and concludes with the message on line \ref{mapc}.5 to indicate that one file was generated. This file is shown in Listing \ref{mapout} and can be used by any programs that consume OpenCL code (e.g. a C program that invokes the generated kernels via the OpenCL host API). 
We will show in the thesis, but omit in this proposal, that for targets with Python bindings, such as OpenCL, CUDA, C, Java or Python itself, generic functions can be executed directly, without any of the explicit compilation steps in Listings \ref{compscript} and \ref{mapc}. This represents a form of staged compilation.  In this setting, the dynamic type of a Python value determines, at the point of invocation, a static type assignment for the argument of an Ace generic function.

\subsubsection{Types}
\begin{codelisting}
\begin{lstlisting}[style=Bash]
> `acec listing2.py`
Hello, compile-time world!
[acec] listing2.cl successfully generated.
\end{lstlisting}
%$ `acec listing3.py`
%Hello, compile-time world!
%[ace] TypeError in listing1.py (line 6, col 28): 
%      'GenericFnType(negate)' does not support [].
%[acec] listing3.cl successfully generated.
\caption{Compiling \texttt{listing\ref{compscript}.py} using the \texttt{acec} compiler.}
\label{mapc}
\end{codelisting}
\begin{codelisting}
\lstinputlisting[style=OpenCL]{../ace-pldi14/listing5.cl}
\caption{[\texttt{listing\ref{compscript}.cl}] The OpenCL file generated by Listing \ref{mapc}.}
\label{mapout}
\end{codelisting}
Lines 7-9 of Listing \ref{compscript} construct the types used to generate concrete functions from the generic function \verb|map| on lines 11 and 12. In Ace, types are themselves values that can be manipulated at compile-time. Python is thus Ace's type-level language. %This stands in contrast to other contemporary languages, where user-defined types (e.g. datatypes, classes, structs) are written declaratively at compile-time but cannot be constructed, inspected or passed around programmatically. 
More specifically, types are instances of a Python class that implements the \verb|ace.ActiveType| interface. Implementing this interface is analagous to defining a new type constructor in @$\lambda$.

As Python values, types can be assigned to variables when convenient (removing the need for  facilities like \verb|typedef| in C or \verb|type| in Haskell). Types, like all compile-time objects derived from Ace base classes, do not have visible state and operate in a referentially transparent manner (by constructor memoization, which we do not detail here).% These types are all implemented in the \verb|clx| library imported on line 1, none are built into Ace itself.

The type named \verb|T1| on line 7 directly implements the logic of the underlying OpenCL type \verb|global float*|: a pointer to a 32-bit floating point number stored in the compute device's global memory (one of four address spaces defined by OpenCL \cite{opencl11}). It is constructed by applying \verb|clx.Ptr|, which is an Ace type constructor corresponding to pointer types, to a value representing the  address space, \verb|clx.global_|, and the type being pointed to. That type, \verb|clx.float|, is in turn the \verb|clx| type corresponding to \verb|float| in OpenCL (which, unlike in C99, is always 32 bits). 
The \verb|clx| library contains a full implementation of the OpenCL type system (including  complexities, like promotions, inherited from C99).
Ace is \emph{unopinionated} about issues like memory safety and the wisdom of such promotions. We will discuss how to implement, as libraries, abstractions that are higher-level than raw pointers, or simpler numeric types, but Ace does not prevent users from choosing a low level of abstraction or ``interesting'' semantics if the need arises (e.g. for compatibility with existing libraries). We also note that we are being more verbose than necessary for the sake of pedagogy. The \verb|clx| library includes more concise shorthand for OpenCL's types: \verb|T1| is equal to \verb|clx.gp(clx.f32)|. %Similarly, the decorators in Listings \ref{map} and \ref{metaprogramming} could have been written \verb|clx.cl_fn|.\todo{move this back there probably}

The type \verb|T2| on line 8 is a pointer to a \emph{complex integer} in global memory. It does not correspond directly to a type in OpenCL, because OpenCL does not include primitive support for complex numbers. Instead, the type constructor \verb|clx.Cplx| defines the necessary logic for typechecking operations on complex numbers and elaborating them to OpenCL. This constructor is parameterized by the numeric type that should be used for the real and imaginary parts, here \verb|clx.int|, which corresponds to 32-bit OpenCL integers. Arithmetic operations with other complex numbers, as well as with plain numeric types (treated as if their imaginary part was zero), are supported. When targeting OpenCL, Ace expressions assigned type \verb|clx.Cplx(clx.int)| are compiled to OpenCL expressions of type \verb|int2|, a  \emph{vector type} of two 32-bit integers (a type that itself is not inherited from C99). This can be observed in several places on lines \ref{mapout}.14-\ref{mapout}.21. This choice is merely an implementation detail that can be kept private to \verb|clx|. An Ace value of type \verb|clx.int2| (that is, an actual OpenCL vector) \emph{cannot} be used when a \verb|clx.Cplx(clx.int)| is expected (and attempting to do so will result in a static type error).

The type \verb|TF| on line 9 is extracted from the generic function \verb|negate| constructed in Listing \ref{compscript}. Generic functions, according to Sec. \ref{genfn}, have not yet had a type assigned to them, so it may seem perplexing that we are nevertheless extracting a type from \verb|negate|. Although a conventional arrow type cannot be assigned to \verb|negate|, we can give it a \emph{singleton type}: a type that simply means ``this expression is the \emph{particular} generic function \verb|negate|''. This type could also have been explicitly written as \verb|ace.GenericFnType(listing2.negate)|. During typechecking and translation of \verb|map_neg_f32| and \verb|map_neg_ci32|, the call to \verb|f| on line 6 of Listing \ref{map} uses the type of the argument to generate a concrete function from the generic function that inhabits the singleton type of \verb|f| (\verb|negate| in both cases shown). This is why there are two versions of \verb|negate| in the output in Listing \ref{mapout}. In other words, types \emph{propagate} into generic functions -- we didn't need to compile \verb|negate| explicitly. In effect, this scheme enables higher-order functions even when targeting languages, like OpenCL, that have no support for higher-order functions (OpenCL, unlike C99, does not support function pointers). Interestingly, because they have a singleton type, they are higher-order but not first-class functions. That is, the type system would prevent you from creating a heterogeneous list of generic functions. Concrete functions, on the other hand, can be given both a singleton type and a true function type. For example, \verb|listing2.negate[[clx.int]]| could be given type \verb|ace.Arrow(clx.int, clx.int)|. The base determines how to convert the Ace arrow type to an arrow type in the target language (e.g. a function pointer for C99, or an integer that indexes into a jump table constructed from knowledge of available functions of the appropriate type in OpenCL).

%Type assignment to generic functions is similar in some ways to template specialization in C++. In effect, both a template header and type parameters at call sites are being generated automatically by Ace.% This simplifies a sophisticated feature of C++ and enables its use with other targets like OpenCL. %Other uses for C++ templates (and the preprocessor) are subsumed by the metaprogramming features discussed above\todo{cite template metaprogramming paper}.

%\subsection{Within-Function Type Resolution}
%\begin{codelisting}
%\lstinputlisting{listing6.py}
%\caption{\texttt{[listing6.py]} A function demonstrating whole-function type inference when multiple values with differing types are assigned to a single identifier, \texttt{y}.}
%\label{inference}
%\end{codelisting}
%On line 5 in the generic \verb|map| function in Listing \ref{map}, the variable \verb|gid| is initialized with the result of calling the OpenCL primitive \verb|get_global_id|.  The type for \verb|gid| is never given explicitly. This is a simple case of Ace's {\em within-function type resolution} strategy (we hesitate to call it \emph{type inference} because it does not, strictly speaking, take a constraint-solving approach). In this case, the type of \verb|gid| will resolve to \verb|size_t| because that is the return type of \verb|get_global_id| (as defined in the OpenCL specification, which the \verb|ace.OpenCL| module follows). The result can be observed on Lines 11 and 24 in Listing \ref{mapout}. 
%
%Inference is not restricted within single assignments, as in the \verb|map| example, however. Multiple assignments to the same identifier with values of differing types, or multiple return statements, can be combined if the types in each case are compatible with one another (e.g. by a subtyping relation or an implicit coercion). In Listing \ref{inference}, the \verb|threshold_scale| function assigns different values to \verb|y| in each branch of the conditional. In the first branch, the value \verb|0| is an \verb|int| literal. However, in the second branch of the loop, the type depends on the types of both arguments, \verb|x| and \verb|scale|. We show two choices for these types on Lines 11 and 12. Type inference correctly combines these two types according to OpenCL's C99-derived rules governing numeric types (defined by the user in the \verb|OpenCL| module, as we will describe in Section \ref{att}). We can verify this programmatically on Lines 12 and 13. Note that this example would also work correctly if the assignments to \verb|y| were replaced with \verb|return| statements (in other words, the return value of a function is treated as an assignable for the purpose of type inference).

\subsection{Remaining Tasks}
\begin{itemize}
\item We need to fix a minor issue raised by a reviewer in the semantics of @$\lambda$ about variables in internal terms when they are substituted.
\item We need to write full proofs of the theorems stated in the @$\lambda$ paper.
\item We need to develop an elaboration of a subset of the Python grammar to @$\lambda$, to connect the two.
\item We need to detail the examples in Sec. 4 of the Ace paper more substantially.
\end{itemize}

### 2014-02-18 thesis-proposal/parsing.tex
Further updates

@@
 \section{Type-Specific Languages}\label{aparsing}
-General-purpose abstraction mechanisms are quite powerful.
-By using a general-purpose abstraction mechanism to encode a data structure, one immediately inherits a body of primitive operations, established reasoning principles, well-optimized implementations and tool support.
+General-purpose abstractions are quite powerful.
+By using a general-purpose abstraction to define a data structure, one immediately benefits from a body of useful operations, established reasoning principles, well-optimized implementations and tool support.
 For example, lists can be encoded using inductive datatypes, the general-purpose abstraction mechanism that most functional programming languages emphasize.
@@
 
-While inheriting these operations and reasoning principles can be quite useful, inheriting the associated general-purpose syntax can sometimes be a liability.
+While these operations and reasoning principles can be quite useful, the associated general-purpose syntax can sometimes be a liability.
 For example, few would claim that writing a list of numbers as a sequence of \li{Cons} cells is convenient:
@@
 
-Because lists are a common data structure, many languages include specialized syntax for introducing them, e.g.
+Because lists are a common data structure, many languages include \emph{literal syntax} for introducing them, e.g.
 \li{[1, 2, 3, 4]}.
@@
 Although list, number and string literals are nearly ubiquitous features of modern languages, some languages also  provide specialized notation for other common data structures (like maps and sets), data formats (like XML and JSON), query languages (like regular expressions and SQL), markup languages (like HTML) and others.
-For example, a language with built-in notation for HTML and SQL, with type-safe interpolation of host language terms within curly braces, might allow:
+For example, a language with built-in notation for HTML and SQL, with type-safe \emph{interpolation} of host language terms within curly braces, might define:
 \begin{lstlisting}
@@
 \end{lstlisting}
-to be shorthand for:
+as shorthand for:
 \begin{lstlisting}
@@
 For example, conventional notations for sets and maps are both delimited by curly braces.
-When Python introduced set literals, it chose to distinguish them based on whether the literal contained only values (e.g.
-\verb|{3}|), or key-value pairs (\verb|{"x": 3}|).
+When Python introduced set literals, it chose to distinguish them based on whether the literal contained only elements (e.g.
+\verb|{3}|), or key-element pairs (\verb|{"x": 3}|).
 But this causes an ambiguity with the syntactic form \verb|{ }| -- should it mean an empty set or an empty map (called a dictionary in Python)?
@@
 So although languages that allow users to introduce new syntax from within libraries appear to hold promise for the reasons described above, providing this form of extensibility is non-trivial because there is no longer a central designer making decisions about such ambiguities.
-In most existing related work, the burden of resolving ambiguities falls to the clients of extensions.
+In most existing related work, the burden of resolving ambiguities falls to the clients using conflicting extensions.
 For example, SugarJ \cite{erdweg2011sugarj} and other extensible languages generated by Sugar* \cite{erdweg2013framework} allow providers to extend the base syntax of the host language with new forms, like set and map literals.
@@
 These kinds of conflicts occur in a variety of circumstances: HTML and XML, different variants of SQL, JSON literals and dictionaries, or simply different implementations (``desugarings'') of the same specialized syntax (e.g.
-two regular expression engines) can all cause problems.
+two regular expression engines) can all cause problems that may be difficult to anticipate.
 
 In this work, we will describe an alternative parsing strategy that avoids these problems by shifting responsibility for parsing certain \emph{generic literal forms} into the typechecker.
-The typechecker, in turn, defers responsibility to user-defined types, by treating the body of the literal as a term of the   \emph{type-specific language (TSL)} associated with the type it is being checked against.
+The typechecker, in turn, defers responsibility to named user-defined types, by treating the body of the literal as a term of the   \emph{type-specific language (TSL)} associated with the type it is being checked against.
 The TSL is responsible for rewriting this term to ultimately use only general-purpose notation.
@@
 We develop our work as a variant of a new programming language being developed by our group called Wyvern \cite{Nistor:2013:WST:2489828.2489830}.
-To allow us to focus on the essence of our proposal, the variant of Wyvern we will describe in this thesis is simpler than the variant previously described: it is purely functional (there are no effects or mutable state) and it does not enforce a uniform access principle for objects (fields can be accessed directly).
-Objects are thus essentially recursive labeled product types with simple methods (functions that require a self-reference).
+To allow us to focus on the essence of our proposal, the variant of Wyvern we will describe in this thesis is simpler than the variant previously described: it is purely functional (there are no effects other than non-termination) and it does not enforce a uniform access principle for objects (fields can be accessed directly).
+Objects are thus essentially records with simple methods (functions that are automatically given a self-reference) for convenience.
 We also add recursive labeled sum types, which we call \emph{case types}, that are quite similar to datatypes in ML.
@@
 let imageBase : URL = new
-  val protocol : URLString = $\text{"}$SUShttpEUS$\texttt{"}$
-  val subdomain : URLString = $\texttt{"}$SUSimagesEUS$\texttt{"}$
+  val protocol = $\text{"}$SUShttpEUS$\texttt{"}$
+  val subdomain = $\texttt{"}$SUSimagesEUS$\texttt{"}$
   (* ... *)
@@
 String interpolation
-never occurs.
-Note that the delimiters that are used to go from Wyvern to a TSL are controlled by Wyvern (those in Fig.
-\ref{f-delims}) while the delimiters that can be used to go from a TSL back to Wyvern are controlled by the TSL.
+never occurs. Note that the delimiters that are used to go from Wyvern to a TSL are controlled by Wyvern (those in Fig. \ref{f-delims}) while the TSL controls how to return to Wyvern.
 
@@
 \ref{f-delims}, and introduced by a form of \emph{forward reference}, written \li{~} (``tilde''), on the previous line.
-Because the forward reference occurs in a position -- as the argument to the function \li{serve} -- where the expected type is \li{HTML}, the literal body is governed by that type's TSL.
+Because the forward reference occurs in a position where the expected type is \li{HTML}, the literal body is governed by that type's TSL.
 The forward reference will be replaced by the general-purpose term, of type \li{HTML}, generated by the TSL during typechecking.
@@
 \ref{f-htmltype}.
-A TSL is associated with a type, forming an \emph{active type}, using a more general mechanism for associating a value with a type.
-A value associated with a type is called the type's  metadata, and is introduced as shown on line 8 of Fig.
+A TSL is associated with a type, forming an \emph{active type}, using a more general mechanism for associating a value with a named type.
+A value associated with a named type is called its \emph{metadata}, and is introduced as shown on line 8 of Fig.
 \ref{f-htmltype}.
 For the purposes of this work, metadata values will always be of type \li{HasTSL}, an object type that declares a single field, \li{parser}, of type \li{Parser}.
-The \li{Parser} type is an object type declaring a single method, \li{parse}, that transforms a \li{TokenStream} to a Wyvern AST, which is a value of type \li{Exp}.
-This is in turn a case type that encodes the syntax of Wyvern expressions.
+The \li{Parser} type is an object type declaring a single method, \li{parse}, that transforms a \li{TokenStream} extracted from a literal body to a Wyvern AST, which is a value of type \li{Exp}.
+This is a case type that encodes the abstract syntax of Wyvern expressions.
 Fig.
-\ref{f-builtins} shows these types.
+\ref{f-builtins} shows portions of the declarations of these types, which live in the Wyvern \emph{prelude} (a collection of ordinary Wyvern types that are automatically loaded before any other).
 
@@
 They support the full Wyvern concrete syntax as well as an additional delimited form, written with \li{%}s, that supports ``unquoting'': interpolating another AST into the one being generated.
-Again, interpolation is type safe and structured, rather than based on string interpolation.
+Again, interpolation is safe and structural, rather than based on string interpolation.
 
@@
 helper functions) in a safe way.
-We formalize this hygiene mechanism by bifurcating the context in our static semantics.
+We formalize this hygiene mechanism by splitting the context in our static semantics.
 \item We provide several examples of TSLs throughout the paper, but to examine how broadly applicable the technique is, we conduct a simple corpus analysis, finding that string languages are used ubiquitously in existing Java code  (collaborative work with Darya Kurilova).

### 2014-02-18 thesis-proposal/problem.tex
Further updates

@@
 Furthermore, even if extensions can be shown not to cause such problems in isolation, there may still be conflicts between extensions that could weaken their semantics, leading to subtle problems that only appear at link-time.
-For example, if two active libraries introduce the same syntactic form but back it with differing (but individually valid) semantics, the issue would only manifest itself when both libraries were imported somewhere within the same compilation unit.
+For example, if two active libraries introduce the same syntactic form but back it with differing (but individually valid) semantics, the issue would only manifest itself when both libraries were imported somewhere within the same scope.
 These kinds of safety issues have plagued previous attempts to design language-integrated extensibility mechanisms.

### 2014-02-21 thesis-proposal/parsing.tex
Essentially done Sec. 4, beginning of Sec. 5, rough draft of end of Sec. 5

@@
   | FromTS of Tokenstream
-  | Literal of TokenStream
   | Error of ErrorMessage
@@
 \item We do not yet have a clean mechanism for preventing unintentional variable \emph{shadowing}, only unintentional variable \emph{capture}.
-It may be possible to prevent shadowing by making it impossible for variables to be introduced into interpolated Wyvern expressions directly (so that function application is the only way to pass data from a TSL to Wyvern code, as in the \li{Parser} TSL above).
-\item In macro systems like Scheme's, free variables in generated ASTs refer to their values within the macro definition, rather than where they are inserted.
+It may be possible to prevent shadowing by preventing variables from leaking into interpolated Wyvern expressions (so that function application is the only way to pass data from a TSL to Wyvern code, as in the \li{Parser} TSL above).
+\item In macro systems like Scheme's, free variables in generated ASTs refer to their bindings within the macro definition, rather than where they are inserted.
 To support this, we have introduced the \li{valAST} operator.
-Using it is a bit verbose and awkward -- every free variable in a quasiquote must be inside \li{valAST}.
-We believe we can insert \li{valAST} automatically from within the definition of the TSL for \li{Exp} using a very simple extension to the interpolation mechanism that allows interpolation of single variables, rather than whole expressions, and plan to write up how this works.
+Using it is a bit verbose and awkward -- every free variable \li{x} in a quasiquote must instead be replaced with an interpolated term of the form \li{valAST(x)} for the resulting \li{Exp} to meaningful as the return value of the \li{parse} function.
+We believe we can insert \li{valAST} automatically from within the definition of the TSL for \li{Exp} using single variable (rather than whole expression) interpolation.
 \end{enumerate}

### 2014-02-26 thesis-proposal/att (Stheno's conflicted copy 2014-02-23).tex
More progress on Sec. 5

--- 2014-02-18 thesis-proposal/att (Stheno's conflicted copy 2014-02-17).tex
@@
 
-\section{Active Type Synthesis}\label{att}
-General-purpose abstraction mechanisms, like those available in Wyvern and other languages today, are powerful, and should be used when possible.
-But relying on a small fixed collection of them may not always be satisfactory, for several reasons:
-\begin{enumerate}
-\item General-purpose abstraction mechanisms themselves are considerably varied.
-Tyconsider, for example, the many different ways languages expose product types: simple products, $n$-ary products, labeled products\footnote{We use the phrase \emph{labeled product} to avoid the connotations of related terms like \emph{record}, \emph{struct} and \emph{object}.}, structurally-typed labeled products, labeled products with mutable fields, labeled products with methods, labeled products with prototypic inheritance, labeled products with delegation, labeled products with class-based inheritance, and various combinations of these.
-New mechanisms and variants are constantly being developed.
-To allow these new general-purpose mechanisms to be evaluated in a controlled manner in practice, it must be possible to import them into existing codebases in a piecemeal fashion.
-\item Specialized type systems that directly enforce stronger invariants than general-purpose abstraction mechanisms are capable of straightforwardly enforcing are often developed.
-In addition to the examples of type systems for regular expressions in Sec.
-\ref{regex} (\cite{a,b}), there are also specialized type systems for parallel programming \cite{a,b}, concurrent programming \cite{cml,a,b}, distributed programming \cite{tom7}, reactive programming \cite{reactiveml}, authenticated data structures \cite{popl13}, security policies \cite{walker00}, information flow \cite{smith2001}, web programming \cite{sandholm00}, aliasing \cite{naden12}, network protocols \cite{sekar99}, units of measure \cite{keneddy} and many others.
-These type systems are implemented by language-external means (e.g.
-by creating a new variant of ML, Java or C), leading to the problems of Sec.
-\ref{external-approaches}.
-\item Interoperability layers with external and legacy systems are another major area where support for directly implementing a new type system (that of the external language) and implementing its primitives in terms of a low-level interoperability API would be useful (e.g.
-\cite{CinJava}).
-%\item General-purpose abstraction mechanisms are typically implemented uniformly.
-For example, inductive datatypes are typically implemented using tagged pointers to a payload.
-In some cases (e.g.
-natural numbers) it may be desirable to implement an abstraction in an alternative manner (e.g.
-as an integer) so that certain operations perform better.
-Parallel abstractions are of particular note, as they can benefit substantially from specialized implementation strategies on specialized hardware (e.g.
-a GPU).
-Note that if control over internal implementation details were the only relevant use case, abstract datatypes might be appropriate.
-\end{enumerate}
+\section{Type-Oriented Mechanisms for Semantic Extensions}\label{att}
 
-In this section, we will focus on language-integrated, type-oriented mechanisms for implementing extensions to the static and dynamic semantics of programming languages.
+In this and the following two sections, we will focus on language-integrated, type-oriented mechanisms for implementing extensions to the static and dynamic semantics of programming languages.
 We will leave the syntax fixed, inverting the setup in the previous section.
-We will begin by taking a  first-principles type-theoretic approach in Sec.
-\ref{foundations}, focusing on safety issues, and continue on in Sec.
-\ref{ace} by designing a full language called Ace.
-We will show how a range of statically-typed general-purpose abstraction mechanisms as well as more specialized abstractions can be expressed as safely-composable extensions within Ace.
-Notably, Ace is itself implemented as a library for Python.
-We discuss how to bootstrap a highly-expressive extensible static type system atop the quotation and object-oriented features available in Python, a uni-typed language.
+We do not believe  the two approaches to be incompatible with one another, but it is useful to consider the issues separately.
 
-\subsection{Foundations}\label{foundations}
-Programming languages are typically organized around a monolithic collection of indexed type and operator constructors.
-Let us begin by considering the simply-typed lambda calculus (STLC).
-It provides a single type constructor, indexed by a pair of types, that we will call $\fvar{Arrow}$ for uniformity.
-It also provides two operator constructors: $\lambda$, indexed by a type, and $\texttt{ap}$, indexed trivially.
-We can specify the static semantics of the STLC as follows, using a uniform abstract syntax where type and operator indices are written in braces to emphasize this way of thinking about its structure:
+\subsection{Foundations}
+Typed programming languages can be considered in terms of the collection of indexed type and operator constructors they provide.
+The simply-typed lambda calculus (STLC), for example, provides a single type constructor, \fvar{Arrow}, indexed by a pair of types, and two operator constructors: $\opvar{lam}$, indexed by a type, and $\opvar{ap}$, which can be thought of as being indexed trivially.
+It's static semantics can be written abstractly like this, to emphasize this uniform way of thinking about type structure:
 \begin{figure}[h]
@@
 }{
-	\jetX{\lambda[\tau]({x}.e)}{\fvar{Arrow}[(\tau, \tau')]}
+	\jetX{\opvar{lam}[\tau]({x}.e)}{\fvar{Arrow}[(\tau, \tau')]}
 }
@@
 }{
-	\jetX{\texttt{ap}[()](e_1; e_2)}{\tau'}
+	\jetX{\opvar{ap}[()](e_1; e_2)}{\tau'}
 }
@@
 
-Although a researcher may casually speak of ``conservatively extending the STLC`` with a primitive natural number type and its corresponding operators (as in G\"odel's T; Figure \ref{nat-statics}) or primitive product types and their  corresponding operators (Figure \ref{pair-statics}), it is clearly not possible to directly introduce these from within the STLC.
-\begin{figure}[h]
-\small
-\vspace{-8pt}
-\begin{mathpar}
-\inferrule[nat-I1]{ }{
-	\jetX{\texttt{z}[()]()}{\fvar{Nat}[()]}
+We might extend this language with universally quantified and recursive types by adding the type constructors $\fvar{Forall}$ and $\fvar{Mu}$, which are indexed by types under a binder, and their associated operator constructors (see \cite{pfpl} for a review).
+It is tempting to stop here: the language now supports general recursion (i.e.
+it is  Turing complete) and the dynamic semantics of sums, products and inductive and co-inductive types can be defined by a method analagous to Church encodings in the untyped lambda calculus \cite{reynolds}\todo{what to cite for this?}.
+This is of clear theoretical and pedagogical interest, but this calculus is plainly impractical as a programming language.
+Reynolds, striking a cautionary tone reminiscent of Perlis in his description of the ``Turing tarpit'' \cite{Perl82a}, put it succinctly \cite{Reynolds94anintroduction}:
+\begin{quote}
+To say that any reasonable function can be expressed by some program is not to say that it can be expressed by the most reasonable program.
+It is clear that the language requires a novel programming style.
+Moreover, it is likely that certain important functions cannot be expressed by their most efficient algorithms.
+\end{quote}
+
+A simple calculus, equipped perhaps with some purely syntactic desugarings as given by a Church-style  weak encoding, can often suffice when only the dynamic semantics of abstractions are of importance.
+But continuing to build new type and operator constructors (collectively, \emph{constructs}) into a language can make statically reasoning about programs simpler and more precise, support a more natural programming style and endow the language with a more favorable cost semantics.
+
+Consistent with this view, typed programming languages generally expose a richer set of constructs externally than their compilers work with internally.  
+A functional language like ML, for example, might expose $n$-ary product types (tuples), labeled product types (records), recursive labeled sum types (datatypes) and forms of type abstraction (polymorphism and abstract  types).
+A compiler for the language might then, after typechecking a program according to the richer static semantics that these constructs define, translate it to an intermediate  language (IL) with only simple product and sum types, and perhaps some machine-oriented constructs like arrays, for further compilation to machine code.
+Correctly implementing the dynamic semantics of the language becomes the primary concern during this phase, so a simple IL that decreases the number of cases that must be considered when implementing and verifying the compiler is a wise choice.
+
+It is again tempting to stop here.
+Indeed, languages like ML are often referred to as ``general-purpose languages'' and have been used successfully for a range of computing tasks.
+The constructs they provide certainly occupy a ``sweet spot'' in the design space, and a variety of higher-level abstractions can be strongly defined in a satisfying manner.
+But claims that they are suitable for \emph{all} purposes are not backed by any ``universality'' theorems stronger than those for the simple calculus described above.
+In fact, situations continue to arise in both research and practice where new constructs are clearly needed because a useful typed abstraction can only be weakly defined in terms of existing constructs, or the definition is impractically verbose or inefficient:
+
+\begin{enumerate}
+\item General-purpose abstractions continue to evolve and admit variants.
+There are many  variations on record types, for example: records with functional record update, prototypic delegation or fixed field ordering all have valid uses, and we will give all of these as examples.
+Similarly, sum types are awkward to encode and often unsafe to define using object-oriented constructs, and also admit variants (e.g.
+open sums, as we will discuss).
+
+\item Perhaps more interestingly, specialized type systems that enforce stronger invariants than general-purpose constructs are capable of enforcing are often proposed by researchers.
+One need not take more than a cursory glance through the literature to discover specialized type systems for parallel programming \cite{a,b}, concurrent programming \cite{cml,a,b}, distributed programming \cite{tom7}, dataflow programming \cite{reactiveml}, authenticated data structures \cite{popl13}, information flow \cite{walker00, smith2001}, databases \cite{db}, aliased references \cite{naden12}, effects \cite{effects},  network protocols \cite{sekar99}, regular expressions \cite{regex}, units of measure \cite{keneddy} and many others.
+These examples are all implemented non-orthogonally (typically as extended versions of an existing language).
+\item Interoperability layers that allow programmers to safely and naturally interact with libraries written in a different language require enforcing the type system of the foreign language within the calling language.
+Using  an interoperability layer that does not enforce these invariants in the calling language can lead to safety issues, even when both languages are separately known to be safe.
+Interoperability is a particularly important concern when transitioning to a new language, so this is likely a significant barrier to adoption of modern language constructs in practice.
+\end{enumerate}
+An extensible programming language could address these problems by providing a safe language-integrated mechanism for introducing new type and operator constructors and implementing their associated static and dynamic semantics directly, and we will explore this in the next two sections.
+Let us first take a detour into a typical compiler to better motivate the mechanisms that we will introduce into the language.
+%Library developers need only consider which abstractions are most appropriate for their domain, without also considering whether these constructs can be exposed using abstractions appropriate to the domains of client code.
+Clients can simply import any necessary constructs when using a library that relies on them, preserving safety and ease-of-use without the use of  wrappers and glue code.
+We show this competing approach in Figure \ref{approaches}(b).
+%Researchers and domain experts thus gain the ability to distribute new ideas for evaluation to a broader development community without requiring the approval of maintainers of mainstream languages, large-scale porting of code or explicit interoperability layers.
+
+\subsection{From Extensible Compilers to Extensible Languages}\label{evolution}
+The monolithic character of most programming languages is reflected in, and perhaps influenced by, the most common abstractions used for implementing programming languages.
+Let us consider an implementation of the STLC. 
+A compiler written using a functional language will invariably represent the primitive type  and operator constructors using {closed} recursive sums. 
+For example, a simple implementation in Standard ML might be based around these datatypes (where variables, binders and operator application to arguments are handled separately, not shown)\footnote{This is essentially the approach that we took in the infrastructure for the coding assignments in 15-312.}:
+\begin{lstlisting}
+  datatype Ty = Arrow of Ty * Ty
+  datatype Op = Lam of Ty | Ap 
+\end{lstlisting}
+
+%The compiler front-end, which consists of a typechecker and translator to a suitable intermediate language,    proceeds by exhaustive case analysis over the constructors of \verb|Op|.
+
+In a class-based object-oriented implementation of the STLC, we might instead encode type and operator constructors as subclasses of abstract classes \verb|Ty| and \verb|Op|.
+If typechecking and translation proceed by the common \emph{visitor pattern}, dispatching against a fixed set of {known} subclasses of \verb|Op|, then we encounter a similar issue: everything is defined at once.
+
+This issue was first discussed by Reynolds \cite{Reynolds75} and was dubbed the \emph{expression problem} by Wadler \cite{wadler1998expression}.
+A number of mechanisms have since been proposed that allow new cases to be added to data types and the functions that operate over them in a modular manner.
+In functional languages, we might turn to \emph{open datatypes and open functions} \cite{conf/ppdp/LohH06, Millstein:2002:MTH:581478.581489}.
+For example, we might add a new type constructor, \verb|Tuple|, indexed by a list of types, and two new operator constructors: an introductory form, \verb|NewTuple|, indexed trivially, and an elimination form, \verb|Prj|, indexed by a natural number, as follows:
+\begin{lstlisting}
+  newcon Tuple of Ty list extends Ty
+  newcon NewTuple extends Op
+  newcon Prj of nat extends Op
+\end{lstlisting}
+
+The logic for functionality like typechecking and translation, if written using open functions, can then be implemented for only these new cases.
+For example, the \verb|optype| function that synthesizes a type for an operator invocation given a context and a list of arguments might be extended to support the new constructor \verb|Prj| like so:
+\begin{lstlisting}
+  optype (ctx, Prj(n), [e]) = case exptype (ctx, e) of 
+      Tuple(ts) => (case List.nth n ts of 
+          Some t => t
+        | _ => raise IndexError("<projection index out of bounds>"))
+    | _ => raise TypeError("<tuple expected>")
+  optype (ctx, Prj(n), _) = raise ArityError("<expected 1 argument>")
+\end{lstlisting}
+
+In a class-based object-oriented language like Java, we might dispense with the visitor pattern and instead invert control to methods of \verb|Op|.
+
+\begin{lstlisting}[language=Java]
+class Prj extends Op {
+  Prj(Nat n) { this.n = n; }
+  Nat n;
+  Ty optype(Context ctx, List<Exp> args) {
+    if (args.length != 1) throw new ArityError("<expected 1 argument>")
+    Ty t = args[0].exptype(ctx);
+    if (t instanceof Tuple) {
+      Tuple t = (Tuple) t;
+      try {
+        return t.types[this.n]
+      } catch OutOfBoundsException(e) {
+        throw new IndexError("<projection index out of bounds>");
       }
+    } else throw new TypeError("<tuple expected>");
+  }
+}
+\end{lstlisting}
 
-\inferrule[nat-I2]{
-	\jetX{e}{\fvar{Nat}[()]}
-}{
-	\jetX{\texttt{s}[()](e)}{\fvar{Nat}[()]}
-}
+If we allowed users to load definitions like these into our compiler, we will have succeeded in creating an  extensible compiler, albeit one where orthogonality is not guaranteed (we will return to this point).
+We have not, however, created an extensible programming language because other compilers for the same language will not necessarily support the same extensions.
+These mechanisms are not \emph{language-integrated}, so if our newly-introduced constructs are exposed at a library's  interface boundary, clients of that library who are using different compilers face the same problems with client compatibility that those using different languages face (as described in Sec.
+\ref{external-approaches}).
+%That is, {extending a language by extending a single compiler for it is morally equivalent to creating a new language}.
+Nevertheless, these two strategies give us the conceptual seeds for the approaches that we will take in Secs. \ref{atlam} and \ref{ace}, respectively. 
+In both cases, we introduce into the language a mechanism similar to, but more constrained and specialized  than, the two above.
+Being part of the language, all compilers must then support all extensions, avoiding the client compatibility issue.\footnote{For  languages where a canonical implementation {is}, unfortunately, the specification (e.g.
+Haskell and Scala), the mechanisms we introduce can be left in the compiler if the build system loads and activates all needed extensions when linking two libraries.
+The advances related to safety, orthogonality and conservativity apply also to this design, but we do not consider it further in this thesis.}
 
-\inferrule[nat-E]{
-	\jetX{e_1}{\fvar{Nat}[()]}\\
-	\jetX{e_2}{\tau}\\
-	\jet{\eCtxX{\eCtxX{\eCtx}{x}{\fvar{Nat}[()]}}{y}{\tau}}{e_3}{\tau}
-}{
-	\jetX{\texttt{natrec}[()](e_1; e_2; x.y.e_3)}{\tau}
-}
-\end{mathpar}
-\vspace{-8pt}
-\caption{Static semantics of natural numbers.}\label{nat-statics}
-\vspace{-15pt}
-\end{figure}
-\begin{figure}[h]
-\small
-\begin{mathpar}
-\inferrule[Prod-I]{
-	\jetX{e_1}{\tau_1}\\
-	\jetX{e_2}{\tau_2}
-}{
-	\jetX{\texttt{pair}[()](e_1; e_2)}{\fvar{Prod}[(\tau_1, \tau_2)]}
-}
+% We will show how, by integrating typed compilation techniques and a form of data hiding with similarities to abstract types into the semantics of operators, we can achieve type safety and conservativity of the language as a whole, despite this level of extensibility.
+Decidability is achieved by ensuring that the type-level language is total.
+Type equality is made decidable in a simple manner: by preventing type-level functions from appearing in type indices.
 
-\inferrule[Prod-E1]{
-	\jetX{e}{\fvar{Prod}[(\tau_1, \tau_2)]}
-}{
-	\jetX{\texttt{prl}[()](e)}{\tau_1}
-}
+%The design described above suggests we may now need to add another layer to our language, above the type-level language (conventionally, $\tau$) and expression language (conventionally, $e$), where extensions are implemented.
+In fact, we will show that \textbf{a natural place for type system extensions is within the type-level language}.
+The intuition  is that extensions to a statically typed language's semantics will need to manipulate types as values at compile-time.
+Many languages already allow users to write type-level functions for various reasons, effectively supporting this notion of types as values at compile-time.
+The type-level language is often constrained by its own type system, where the types of type-level terms are called \emph{kinds} for clarity,  that prevents type-level functions from causing problems during compilation.
+The kind system is often more constraining than the type system is (e.g.
+it might permit only terminating functions, to ensure that typechecking is decidable).
+This is precisely the structure that a distinct extension layer would have, and so we will show that is quite natural to unify the two.
 
-\inferrule[Prod-E2]{
-	\jetX{e}{\fvar{Prod}[(\tau_1, \tau_2)]}
-}{
-	\jetX{\texttt{prr}[()](e)}{\tau_1}
-}
-\end{mathpar}
-\vspace{-8pt}
-\caption{Static semantics of products}\label{pair-statics}
-\end{figure}
+We begin in Sec.
+\ref{atlam} by developing a core calculus called @$\lambda$ and focusing on theoretical issues, including type safety, decidability and {conservativity} (that extensions cannot weaken one another, in any combination).
+We will also sketch a flexible type-directed desugaring mechanism that targets this core calculus and give examples of simple but interesting extensions that are only weakly definable using the general-purpose constructs in functional languages.
+We then continue on in Sec.
+\ref{ace} by designing a full programming language called Ace.
+We will show richer examples of statically-typed general-purpose abstractions, as well as more specialized abstractions, that can be expressed as extensions within Ace.
+Ace  uses Python as its type-level language and supports the definition of user-defined internal languages, rather than picking a fixed target as in @$\lambda$.
+Using a type-level language that provides few strictly enforced safety guarantees makes giving strong theoretical results about Ace more difficult, but we include checks similar to those in @$\lambda$ that require intention to subvert.
+The claimed contribution is thus fundamentally that Ace shows how to {bootstrap} a highly-expressive extensible static type system atop facilities, like quotations and objects, that are already available in widely-adopted languages.
+We argue that this is a useful bridge between existing monolithic languages and extensible languages.
+We conjecture that it would be possible to bootstrap a statically-typed subset of Python that, if used ubiquitously during compilation, would be sufficient to provide stronger theoretical guarantees as well.
 
-One recourse researchers have in such situations is to attempt to encode new constructs in terms of existing constructs.
+%But, as mentioned in Section \ref{language-integrated-approaches}, some significant challenges must be addressed before such a mechanism can be relied upon.
+The desire for expressiveness must be balanced against  concerns about maintaining various safety properties in the presence of arbitrary combinations of user-defined  extensions to the language's core semantics.
+The mechanism must ensure that desirable \emph{metatheoretic properties} (e.g.
+type safety, decidability) of the language are maintained by extensions.
+Because multiple independently developed extensions might be used within one program, the mechanism must further guarantee that extensions are \emph{conservative}.
+These are the issues we seek to address in the next two sections.
+
+\section{@$\lambda$}\label{atlam}
+%Programming languages are typically organized around a monolithic collection of indexed type and operator constructors.
+Let us begin by considering the simply-typed lambda calculus (STLC).
+It provides a single type constructor, indexed by a pair of types, that we will call $\fvar{Arrow}$ for uniformity.
+It also provides two operator constructors: $\lambda$, indexed by a type, and $\texttt{ap}$, indexed trivially.
+We can specify the static semantics of the STLC as follows, using a uniform abstract syntax where type and operator indices are written in braces to emphasize this way of thinking about its structure:
+%
+%Although a researcher may casually speak of ``conservatively extending the STLC`` with a primitive natural number type and its corresponding operators (as in G\"odel's T; Figure \ref{nat-statics}) or primitive product types and their  corresponding operators (Figure \ref{pair-statics}), it is clearly not possible to directly introduce these from within the STLC.
+%\begin{figure}[h]
+%\small
+%\vspace{-8pt}
+%\begin{mathpar}
+%\inferrule[nat-I1]{ }{
+%	\jetX{\texttt{z}[()]()}{\fvar{Nat}[()]}
+%}
+%
+%\inferrule[nat-I2]{
+%	\jetX{e}{\fvar{Nat}[()]}
+%}{
+%	\jetX{\texttt{s}[()](e)}{\fvar{Nat}[()]}
+%}
+%
+%\inferrule[nat-E]{
+%	\jetX{e_1}{\fvar{Nat}[()]}\\
+%	\jetX{e_2}{\tau}\\
+%	\jet{\eCtxX{\eCtxX{\eCtx}{x}{\fvar{Nat}[()]}}{y}{\tau}}{e_3}{\tau}
+%}{
+%	\jetX{\texttt{natrec}[()](e_1; e_2; x.y.e_3)}{\tau}
+%}
+%\end{mathpar}
+%\vspace{-8pt}
+%\caption{Static semantics of natural numbers.}\label{nat-statics}
+%\vspace{-15pt}
+%\end{figure}
+%\begin{figure}[h]
+%\small
+%\begin{mathpar}
+%\inferrule[Prod-I]{
+%	\jetX{e_1}{\tau_1}\\
+%	\jetX{e_2}{\tau_2}
+%}{
+%	\jetX{\texttt{pair}[()](e_1; e_2)}{\fvar{Prod}[(\tau_1, \tau_2)]}
+%}
+%
+%\inferrule[Prod-E1]{
+%	\jetX{e}{\fvar{Prod}[(\tau_1, \tau_2)]}
+%}{
+%	\jetX{\texttt{prl}[()](e)}{\tau_1}
+%}
+%
+%\inferrule[Prod-E2]{
+%	\jetX{e}{\fvar{Prod}[(\tau_1, \tau_2)]}
+%}{
+%	\jetX{\texttt{prr}[()](e)}{\tau_1}
+%}
+%\end{mathpar}
+%\vspace{-8pt}
+%\caption{Static semantics of products}\label{pair-statics}
+%\end{figure}
+%
+%One recourse researchers have in such situations is to attempt to encode new constructs in terms of existing constructs.
 Collections of such encodings are sometimes called \emph{embedded domain-specific languages (EDSLs)} \cite{fowler2010domain}.
@@
 Even when it possible, it may be impractical to work with and optimize.
-For our example of adding natural numbers or products to the STLC, a reasonable encoding is impossible.
+For our example of adding natural numbers or products to the STLC, an encoding is impossible.
 In closely-related languages, like the polymorphic lambda calculus (a.k.a.
@@
 Reynolds, echoing Perlis   \cite{perlisPearl}, puts it succinctly \cite{reynoldsIntroToPolymorphism}:
-
-\begin{quote}
-To say that any reasonable function can be expressed by some program is not to say that it can be expressed by the most reasonable program.
+%
+%\begin{quote}
+%To say that any reasonable function can be expressed by some program is not to say that it can be expressed by the most reasonable program.
 It is clear that the language requires a novel programming style.
-Moreover, it is likely that certain important functions cannot be expressed by their most
-efficient algorithms.
-\end{quote}
+%Moreover, it is likely that certain important functions cannot be expressed by their most
+%efficient algorithms.
+%\end{quote}
+%
 
@@
 
-An extensible programming language could address these problems by providing a language-integrated mechanism for introducing new type and operator constructors and implementing their associated static and dynamic semantics directly.
-%Library developers need only consider which abstractions are most appropriate for their domain, without also considering whether these constructs can be exposed using abstractions appropriate to the domains of client code.
-Clients can simply import any necessary constructs when using a library that relies on them, preserving safety and ease-of-use without the use of  wrappers and glue code.
-We show this competing approach in Figure \ref{approaches}(b).
-%Researchers and domain experts thus gain the ability to distribute new ideas for evaluation to a broader development community without requiring the approval of maintainers of mainstream languages, large-scale porting of code or explicit interoperability layers.
-But, as mentioned in Section \ref{language-integrated-approaches}, some significant challenges must be addressed before such a mechanism can be relied upon.
-The desire for expressiveness must be balanced against  concerns about maintaining various safety properties in the presence of arbitrary combinations of user-defined  extensions to the language's core semantics.
-The mechanism must ensure that desirable \emph{metatheoretic properties} (e.g.
-type safety, decidability) of the language are maintained by extensions.
-Because multiple independently developed extensions might be used within one program, the mechanism must further guarantee their \emph{non-interference}.
-These are the issues we seek to address in this work.
 %\subsection{Theory}\label{atlam}
@@
 
-\subsubsection{From Extensible Compilers to Extensible Languages}\label{evolution}
-The monolithic character of most programming languages is reflected in, and perhaps influenced by, the most common mechanisms used for implementing programming languages.
-Let us consider an implementation of the STLC. 
-A compiler written using a functional language will invariably represent the primitive type  and operator constructors using {closed} recursive sums. 
-A simple implementation in Standard ML could be based around these datatypes, for example:
-\begin{lstlisting}
-  datatype Type = Arrow of Type * Type
-  datatype Exp = Var of var | Lam of var * Type * Exp | Ap of Exp * Exp 
-\end{lstlisting}
-
-The compiler front-end, which consists of a typechecker and translator to a suitable intermediate language (for subsequent compilation by some suitable back-end), will proceed by exhaustive case analysis over the constructors of \lstinline{Exp}.
-
-In a class-based object-oriented implementation of Godel's T, we might instead encode type and operator constructors as subclasses of abstract classes \lstinline{Type} and \lstinline{Exp}.
-Typechecking and translation, however, will proceed by the ubiquitous \emph{visitor pattern}, dispatching against a fixed collection of {known} subclasses of \lstinline{Exp}.
-
-In either case, we encounter the same basic issue: there is no way to modularly add new primitive type and operator constructors to \li{Type} and \li{Exp}, respectively, and provide implementations of their associated typechecking and translation logic.
-Everything is defined in precisely one place.
-%This issue is related to the widely-discussed \emph{expression problem} (in a restricted sense -- we do not consider adding new functions beyond typechecking and translation here, only adding logic to these) \cite{wadler-expression}.
-
-A number of language mechanisms have been proposed that allow new cases to be added to datatypes and the functions that operate over them in a modular manner. 
-In functional languages, we might use \emph{open datatypes} \cite{opendatatypes}\todo{cite}.
-For example, if we wish to extend our implementation of the STLC with product types and we have written our compiler in a language supporting open datatypes, it might be possible to add new constructors:
-\begin{lstlisting}
-  newcon Prod of Type * Type extends Type
-  newcon Pair of Exp * Exp extends Exp
-  newcon PrL of Exp extends Exp
-  newcon PrR of Exp extends Exp
-\end{lstlisting}
-
-The logic for functionality like typechecking and translation can then be implemented for only these new cases.
-For example, the \lstinline{typeof} function that synthesizes a type for an expression given a context could be extended to support the new case \li{PrL} like so:
-\begin{lstlisting}
-  typeof (ctx, PrL(e)) = case typeof (ctx, e) of 
-      Prod(t1, _) => t1 
-    | _ => raise TypeError("<appropriate error message>")
-\end{lstlisting}
-
-If we allowed users to define new modules containing definitions like these and link them into our compiler, we will have succeeded in creating an externally-extensible compiler, albeit one where safety and non-interference is not guaranteed (we will return to this point shortly).
-We have not, however, created an extensible programming language because other compilers for the same language will not necessarily support the same extensions.
-If our newly-introduced constructs are exposed at a library's  interface boundary, clients of that library who are using different compilers face the same problems with client compatibility that those using different languages face (as described in Sec.
-\ref{external-approaches}).
-That is, {extending a language by extending a single compiler for it is morally equivalent to creating a new language}.
-
-We argue that a more appropriate and useful place for extensions like this is directly within libraries, alongside abstractions that can be implemented in terms of existing primitive abstractions.
-To enable this, the language must allow for the introduction of new type constructors, like \lstinline{Prod}, operator constructors, like \lstinline{Pair}, \lstinline{PrL} and \lstinline{PrR}, and their associated type synthesis and translation logic.
-Because this mechanism is integrated into the language specification, all compilers must support it.
-
-The design described above suggests we may now need to add another layer to our language, above the type-level language (conventionally, $\tau$) and expression language (conventionally, $e$), where extensions are implemented.
-In fact, we will show that \textbf{a natural place for type system extensions is within the type-level language}.
-The intuition  is that extensions to a statically typed language's semantics will need to manipulate types as values at compile-time.
-Many languages already allow users to write type-level functions for various reasons, effectively supporting this notion of types as values at compile-time.
-The type-level language is often constrained by its own type system, where the types of type-level terms are called \emph{kinds} for clarity,  that prevents type-level functions from causing problems during compilation.
-The kind system is often more constraining than the type system is (e.g.
-it might permit only terminating functions, to ensure that typechecking is decidable).
-This is precisely the structure that a distinct extension layer would have, and so we will show that is quite natural to unify the two.
-
-\subsubsection{@$\lambda$: Conservatively Extending a Type System From Within}
+%\subsubsection{@$\lambda$: Conservatively Extending a Type System From Within}
 \begin{figure}[t]
@@
 \textbf{programs} & \rho & ::= & \pfam{\familyDf}{\progsort} \pipe \pdef{t}{\kappa}{\tau}{\progsort} \pipe e\\
-\text{operators}		&	\theta	&	::= &	\tops{op}{\kappaidx}{i}{a}{\taut{def}} \pipe 
+		&	\theta	&	::= &	\tops{op}{\kappaidx}{i}{a}{\taut{def}} \pipe 
 												\topp{\theta}{\theta}\\
@@
 \vspace{-10pt}
-\caption{\small Abstract syntax of \atlam.
-Here, $x$ ranges over external and internal language variables, $\tvar{t}$ ranges over type-level variables, $\fvar{Tycon}$ ranges over type constructor names and $\opvar{op}$ ranges over operator constructor names.
-The form $\tstr{str}$ represents string literals, $\iintlit$ represents integer literals and $\oplus$ represents standard total binary operations over integers.%The productions related to the internal language are written using generators $\mathcal{G}$ and $\mathcal{S}$ to avoid duplicating the syntax of common terms.
+\caption{\small Syntax of \atlam.
+Here, $x$ ranges over external and internal language variables, $\tvar{t}$ ranges over type-level variables, $\fvar{Tycon}$ ranges over type constructor names, $\opvar{op}$ ranges over operator constructor names, $\tstr{str}$ ranges over string literals, $\iintlit$ ranges over integer literals and $\oplus$ ranges over the standard total binary operations on integers.%The productions related to the internal language are written using generators $\mathcal{G}$ and $\mathcal{S}$ to avoid duplicating the syntax of common terms.
 \label{grammar}}
@@
 \end{figure}
-We will begin by developing an ``actively-typed'' version of the simply-typed lambda calculus with type-level computation called @$\lambda$.
-In the semantics for this calculus, we will integrate typed compilation techniques and a form of type abstraction to allow us to prove key type safety and non-interference theorems.
-The abstract syntax of @$\lambda$ is in Fig.
+In this section, we will develop an ``actively typed'' version of the simply-typed lambda calculus with simply-kinded type-level computation called @$\lambda$.
+In the semantics for this calculus, we will integrate typed compilation techniques and a form of type abstraction to allow us to prove key type safety, decidability and conservativity theorems.
+The syntax of @$\lambda$ is given in Fig.
 \ref{grammar}.
+An example of a program, which we will review to introduce the remainder of the calculus, is given in Fig.
+\ref{nat-atlam}\todo{use proper syntax}.
+It consists of:
+a type constructor declaration, $\fvar{Nat}$, together with three operator constructors that implement the standard introductory forms for natural numbers as well as the recursive elimination form (as in G\"odel's T \cite{pfpl}).
+Following the type constructor declaration, we apply the constructor trivially to form the type $\tvar{nat}$.
+Finally, we write an external term that uses the operators associated with that constructor and the built-in constructor $\fvar{Parr}$, governing partial functions (due to the presence of the fixpoint operator in the external language), to define an addition function and compute the addition of two and two.
 
-The semantics of the language are in the style of the Harper-Stone elaboration semantics for Standard ML: all terms in an external language, $e$, are simultaneously assigned a type and an elaboration to a term, $\iota$, in a fixed typed internal language \cite{HarperStone, Stone}.
+\paragraph{Overview} A program, $\rho$, consists of a series of static declarations followed by an external term, $e$.
+The syntax for external terms includes variables, functions, fixpoints and a uniform operator invocation form for user-defined operators.
+As indicated, user-defined operators in @$\lambda$ are associated with type constructors (for reasons we will discuss), so the form $\eop{Tycon}{op}{\tauidx}{\splat{e}{1}{n}}$ applies the operator $\opvar{op}$ associated with type constructor $\fvar{Tycon}$ to operator index $\tauidx$ and zero or more arguments, written in our syntax and semantics using the shorthand ${\splat{e}{1}{n}}$ for concision.
+
+There are two forms of static declarations: 1) \textsf{tycon} declares a new type constructor named \fvar{Tycon}, with \emph{index kind} $\kappaidx$, \emph{representation schema} $\taurep$ and \emph{operator declarations} $\theta$ (these will be explained below), and 2) \textsf{def} binds a type-level value of kind $\kappa$ to a type-level variable, $\tvar{t}$, in the remainder of the program (serving as a type-level analog to \textsf{let}).
+This can be used to, for example, give a short name to a type, as in our example.
+A type is a type-level value of kind $\kTypeBlur$ and is introduced by naming a type constructor in scope and providing an index of the appropriate kind: $\ttype{Tycon}{\tauidx}$.
+Types statically classify external terms, as in other simply-typed languages.
+
+Compilation of a program consists of first \emph{kind checking} its static declarations and type-level terms, $\tau$, then type checking the external term and, simultaneously, deriving a translation.
+External terms are given a dynamic semantics by translation to a typed internal language with terms $\iota$ and types $\sigma$.
+In our calculus, the internal language is essentially PCF with simple products, sums and a single base type of integers.
+The forms in the internal language containing type-level terms will be explained below; they are erased during the typechecking process.
+%Kind checking  can be expressed judgementally as follows:\todo{compilation judgement}
+
+The type-level language also includes other constructs: type-level functions, lists, numbers, strings and products are lifted into the type-level language in a completely standard way.
+There is an elimination form for types that allows one to extract its index knowing which constructor produced it.
+
+\paragraph{Type Constructors} Type constructors are indexed by A type constructor is declared by providing a name, like $\fvar{Tuple}$, an index kind, like %$\list$
+
+Note that this calculus be compared to the ``core language'' of expressions and types in a language in Standard ML; a module language similar to ML's could be layered on top of this core in a completely standard way.
+Note also that we do not include polymorphism and recursive types in this calculus (nor can they be defined, because we do not support type binders).
+This is both to simplify the calculus and because it is known that polymorphism is conservative over simple types \cite{breazu1990polymorphism} (by similar arguments, recursive types are conservative over a simply typed language with fixpoints, which we do include).
+
+The semantics of the language are in the style of the Harper-Stone elaboration semantics for Standard ML: all terms in an external language, $e$, are simultaneously assigned a type and an elaboration to a term, $\iota$, in a fixed typed internal language.
 The mechanism checks that any generated elaborations are \emph{type preserving} -- that all external terms of a particular type, $\tau$, elaborate to internal terms with a consistent internal type, $\sigma$ -- following the approach taken in the TIL compiler for Standard ML by Morrisett et al.
@@
 This allows the semantics to be reasoned about compositionally and, when combined with type safety of the internal language, gives us type safety of the extensible external language.
+
 This design ensures that global properties that the internal language maintains are guaranteed to hold no matter which extensions are introduced.
@@
 
-It is reasonable, if we wish to actually implement this language, to permit only the addition of deterministic, decidable rules to the semantics.
-Extension logic is thus implemented in a functional style  (essentially, lifting fragments of the typechecker into the type-level language), rather than extracted from inductive specifications, like those in the Sec.
-\ref{foundations}\footnote{Though we will not explore this further in this thesis, a system for extracting such a function from a declarative specification could potentially be implemented using the techniques in Sec.
+It is reasonable, if we wish to actually program with such a language, to permit only the addition of deterministic, decidable rules to the semantics.
+Extension logic is thus implemented in a functional style  (essentially, lifting fragments of the typechecker into the type-level language), rather than extracted from inductive specifications, like those at the beginning of Sec.
+\ref{att}\footnote{Though we will not explore this further in this thesis, a system for extracting such a function from a declarative specification could potentially be implemented using the techniques in Sec.
 \ref{aparsing} -- a \emph{kind-specific language}!}.
@@
 The type synthesis and translation logic that implements the desired static and dynamic semantics, respectively, for these is implemented in the type-level language.
+
+To enable this, the language must allow for the introduction of new type constructors, like \verb|Tuple|, operator constructors, like \verb|NewTuple| and \verb|Prj|, and their associated static and dynamic semantics.
+Type and operator constructors can no longer be indexed by compiler-internal values in such a design.
+%We are also be considering only simply-typed languages -- those with a clear phase separation between compile-time and run-time -- so we will not consider dependent types, though extensibility mechanisms for dependently typed languages are an interesting avenue of future work.\footnote{Note that types indexed by expressions are distinct from type names that have metadata expressions associated with them, as in Wyvern.
+Metadata does not impact type equality, for example.}
+We will instead use type-level values.
+This requires supporting more kinds of values in the type-level language, conventionally written $\tau$, than just types, and thus the type-level language must have a non-trivial semantics.
+The ``types'' of type-level terms are called \emph{kinds} for clarity.
+Types simply become a particular kind of type-level value (we will call it $\star$) and are introduced uniformly by naming an available type constructor and providing an index of the kind that constructor specifies.
+For example, the type constructor \verb|Tuple| is indexed by type-level values of kind \verb|list[|$\star$\verb|]| so a 3-tuple type might be written \verb|Tuple[cons(nat, cons(string, cons(nat, nil)))]|.
+Note that we do not identify type constructors and type-level functions, as in some kind systems (\verb|Tuple| is \emph{not} itself of kind \verb|list[|$\star$\verb|] |$\rightarrow$\verb| |$\star$).
+One can wrap constructor application with a type-level function to achieve the same effect (and avoid an awkward canonical forms lemma and beta-reduction rules).
+
+Operator constructors are similarly indexed by type-level values.
+For example, the projection operator written \verb|#2| in a language like Standard ML might be written \verb|prj[2]| in our work.
+Here, \verb|prj| is an operator constructor indexed by a type-level natural number.
+To define an operator, we must also provide it with a static and dynamic semantics.
+We will, as in our work on Wyvern and consistent with our exposition above, take an approach similar to Harper and Stone in their definition of Standard ML where the dynamic semantics of external language terms are given by translation into a  typed internal language \cite{Harper00atype-theoretic}.
+Operator definitions must work with types, type indices and operator indices, all of which are type-level values, so we will argue that it is sensible to give these elaborations as type-level functions (appropriately constrained, as we will show).
+
+-----
 
@@
 where semantic equality coincides with syntactic equality).
-This means that types themselves, as well as type-level data structures containing values of equality kind, can be used, while type-level functions cannot.
+This means that types themselves, as well as type-level data structures containing values of kinds with this property, can be used, while type-level functions cannot.
+We leave a richer consideration of type equality to future work for now.
 
 To ensure that compilation is type-preserving, as described above, each type must have an internal type, called its \emph{representation}, associated with it.
-The representation of a type might depend on its index, so each type constructor must declare a type-level function that returns an internal type given an index, called its \emph{representation schema}.
-An internal type, $\sigma$, is reflected into a type-level terms of kind $\kIType$ via the introductory form $\blacktriangledown(\sigma)$.
+The representation of a type might depend on its index, so each type constructor must declare a type-level function, called its \emph{representation schema}, that returns an internal type given a type's index.
+An internal type, $\sigma$, is reflected as a type-level term of kind $\kIType$ via the introductory form $\blacktriangledown(\sigma)$.
 Because $\fvar{Nat}$ is indexed trivially, it simply returns the reflected form of the internal type $\mathbb{Z}$.
@@
 
+--
+
 In the external language, $e$, all operators (other than $\lambda$, for reasons we will discuss) are invoked uniformly.
@@
 In our example, all the operators are indexed trivially, but we will see examples later where non-trivial operator indices are useful.
-We will discuss why operator constructors are associated directly with type constructors, rather than existing as standalone entities, when we discuss the scoping mechanism for the representation schema, which is essential for guaranteeing non-interference.
-This coupling also forms the foundation for a higher-level elaboration mechanism that allows us to use more conventional and concise (though still fixed) syntactic idioms, which we intend to explore in this thesis, but have not yet formalized.
+For the sake of uniformity, we include trivial indices, but in a concrete syntax for such a language, these could be omitted.
+We will discuss why operator constructors are associated directly with type constructors, rather than existing as standalone entities, when we discuss the scoping mechanism for the representation schema.
+This coupling also forms the foundation for the higher-level elaboration mechanism that allows us to use more conventional and concise (though still fixed) syntactic idioms, which we intend to explore in this thesis, but have not yet formalized.
 For now, we simply note that intuitively, most operators are either introduction or elimination forms for a particular type, so we are simply encoding a common idiom.
+Free-floating operator constructors could be introduced to this calculus with little difficulty.
 
@@
 
-\subsection{Ace}\label{ace}
+\section{Ace}\label{ace}
 \lstset{
@@
 
+\subsubsection{Extensibility}
+The core of Ace consists of about 1500 lines of Python code implementing its primary concepts: generic functions, concrete functions, active types, active bases and active targets.
+The latter three comprise Ace's extension mechanism.
+Extensions provide semantics to, and govern the compilation of, Ace functions, rather than logic in Ace's core.
+%Indeed, the name ``Ace'' might itself be an acronym: it is an ``active compilation environment''.
+
+Active types are the primary means for extending Ace with new abstractions.
+An active type, as mentioned previously, is an instance of a class implementing the \verb|ace.ActiveType| interface.
+Listing \ref{cplx} shows an example of such a class: the \verb|clx.Cplx| class used in Listing \ref{compscript}, which implements the logic of complex numbers.
+The constructor takes as a parameter any numeric type in \verb|clx| (line \ref{cplx}.2).
+
+\subsubsection{Dispatch Protocol}
+\begin{codelisting}
+\lstinputlisting{../ace-pldi14/listing9.py}
+\caption{\texttt{[}in \texttt{examples/clx.py]} The active type family \texttt{Ptr} implements the semantics of OpenCL pointer types.}
+\label{cplx}
+\end{codelisting}
+In a compiler for a monolithic language, there would be a \emph{syntax-directed} protocol governing typechecking and translation.
+In a compiler written in a functional language, for example, one would declare datatypes that captured all forms of types and expressions, and the typechecker would perform exhaustive case analysis over the expression forms.
+That is, all the semantics are implemented in one place.
+The visitor pattern typically used in object-oriented languages implements essentially the same protocol.
+This does not work for an extensible language because new cases and logic need to be added \emph{modularly}, in a safely composable manner.
+
+Instead of taking a syntax-directed approach, the Ace compiler's typechecking and translation phases take a \emph{type-directed approach}.
+When encountering a compound term (e.g.
+\verb|e[e1]|), the compiler defers control over typechecking and translation to the active type of a designated subexpression (e.g.
+\verb|e|) determined by Ace's fixed \emph{dispatch protocol}.
+Below are examples of the choices made in Ace.
+\begin{itemize}
+\item Responsibility over {\bf attribute access} (\texttt{e.attr}), {\bf subscripting} (\texttt{e[e1]}),  \textbf{calls} (\verb|e(e1, ..., en)|) and {\bf unary operations} (e.g.
+\verb|-e|) is handed to the type recursively assigned to \texttt{e}.
+\item Responsibility over {\bf binary operations} (e.g.
+\verb|e1 + e2|) is first handed to the type assigned to the left operand.
+If it indicates a type error, the type assigned to the right operand is handed responsibility, via a different method call.
+{Note that this operates like the corresponding rule in Python's \emph{dynamic} operator overloading mechanism.}
+\item Responsibility over \textbf{constructor calls} (\verb|[t](e1, ..., en)|), where \verb|t| is a \emph{compile-time Python expression} evaluating to an active type, is handed to that type, by using Python's eval functionality to evaluate t.
+If \verb|t| evaluates to a type constructor, like \verb|clx.Cplx|, the type is first generated via a class method, as discussed below.
+%\item Responsibility over {\bf simple assignment statements}, is handed to the type of the variable on the left (which, as we will see below, is determined by the active base).
+If this type does not provide special assignment semantics, the base must handle it.%Destructuring assignment is also supported by a somewhat more complex protocol.\todo{clarify}
+\end{itemize}
+
+\subsubsection{Typechecking}
+When typechecking a compound expression or statement, the Ace compiler temporarily hands control to the object selected by the dispatch protocol by calling the method \verb|type_|$X$, where $X$ is  the name of the syntactic form, taken from the Python grammar \cite{python} (appended with a suffix in some cases).
+
+For example, if \verb|c| is a complex number, then the operations \verb|c.ni| and \verb|c.i| extract its non-imaginary and imaginary components, respectively.
+These expressions are of the form \verb|Attribute|, so the typechecker calls \verb|type_Attribute| (line 7).
+This method receives the compilation context, \verb|context|, and the abstract syntax tree of the expression, \verb|node|, and must return a type assignment for it, or raise an \verb|ace.TypeError| if there is an error.
+In this case, a type assignment is possible if the attribute name is either \verb|"ni"| or \verb|"i"|, and an error is raised otherwise (lines 8-10).
+We note that error messages are an important and sometimes overlooked facet of {ease-of-use} \cite{marceau2011measuring}.
+A common frustration with using general-purpose abstraction mechanisms to encode an abstraction is that they can produce  verbose and cryptic error messages that reflect the implementation details instead of the semantics.
+Ace supports custom error messages.%Active types permit direct encoding of semantics and customization of error messages.
+
+Complex numbers also support binary arithmetic operations partnered with both other complex numbers and with non-complex numbers, treating them as if their imaginary component is zero.
+The typechecking rules for this logic is implemented on lines 17-29.
+Arithmetic operations are usually symmetric, so the dispatch protocol checks the types of both subexpressions for support.
+To ensure that the semantics remain deterministic in the case that both types support the binary operation, Ace asks the left first (via \verb|type_BinOp_left|), asking the right (via \verb|type_BinOp_right|) only if the left indicates an error.
+In either position, our implementation begins by recursively assigning a type to the other operand in the current context via the \verb|context.type| method (line 24).
+If supported, it applies the C99 rules for arithmetic operations to determine the resulting type (via \verb|c99_binop_t|, not shown).
+
+Finally, a complex number can be constructed inside an Ace function using Ace's special constructor form: \verb|[clx.Cplx](3,4)| represents $3+4i$, for example.
+The term within the braces is evaluated at \emph{compile-time}.
+Because \verb|clx.Cplx| evaluates not to an active type, but to a class, this form is assigned a type by handing control to the class object via the \emph{class method} \verb|type_New|.
+It operates as expected, extracting the types of the two arguments to construct an appropriate complex number type (lines 50-57), raising a type error if the arguments cannot be promoted to a common type according to the rules of C99 or if two arguments were not provided.% (an exercise for the reader: modify this method to also allow a single argument for when the imaginary part is 0).
+
+%Note that in each example, subexpressions are not \emph{automatically} assigned types.
+Types must be requested from the context.
+This is useful because it allows types to reinterpret subexpressions of terms they control, to supporting more complex constructs (e.g.
+pattern matching on functional datatypes, see Sec.
+\ref{datatypes}).
+
+\subsubsection{Translation}
+Once typechecking a method is complete, the compiler enters the translation phase, where terms in the target language are generated from Ace terms.
+Terms in the target language are generated by calling methods of the \emph{active target} governing the function being compiled.
+The translation phase operates similarly to typechecking, using the dispatch protocol to invoke methods named \verb|trans_|$X$.
+These methods have access to the context and node just as during typechecking, as well as the active target (named \verb|target| here).
+
+As seen in Listing \ref{mapout}, we are implementing complex numbers internally using OpenCL vector types, like \verb|int2|.
+Let us look first at \verb|trans_New| on lines 54-60, where new complex numbers are translated to vector literals by invoking \verb|target.VecLit|.
+This will ultimately  generate the necessary OpenCL code, as a string, to complete compilation (these strings are not directly manipulated by extensions, however, to avoid problems with, e.g.
+precedence).
+For it to be possible to reason compositionally about the correctness of compilation, all complex numbers must translate to terms in the target language that have a consistent target type.
+The \verb|trans_type| method of the \verb|ace.ActiveType| associates a type in the target language, here a vector type like \verb|int2|, with the active type.
+Ace supports a mode where this \emph{representational consistency} is dynamically checked during compilation (requiring that the active target know how to assign types to terms in the target language, which can be done  for our OpenCL target as of this writing).
+
+The translation methods for attributes (line 12) and binary operations  (line 31) proceed in a straightforward manner.
+The context provides a method, \verb|trans|, for recursively determining the translation of subexpressions as needed.
+Of note is that the translation methods can assume that typechecking succeeded.
+For example, the implementation of \verb|trans_Attribute| assumes that if \verb|node.attr| is not \verb|'ni'| then it must have been \verb|'i'| on line 14, consistent with the implementation of \verb|type_Attribute| above it.
+Typechecking and translation are separated into two methods to emphasize that typechecking is not target-dependent, and to allow for more advanced uses, like type refinements and hypothetical typing judgements, that we do not describe here.
+
+\subsubsection{Active Bases}\label{abases}
+Each generic function is associated with an active base, which is an object implementing the \verb|ace.ActiveBase| interface.
+The active base specifies the \emph{base semantics} of that function.
+It controls the semantics of statements and expressions that do not have a clear ``primary'' subexpression for the dispatch protocol to defer to.
+A base is handed control over typechecking of statements and expressions in the same way as active types: via \verb|type_|$X$ and \verb|trans_|$X$ methods.
+Each function can have a different base.
+
+Literals are the most prominent form given their semantics by an active base.
+Our example active base, \verb|clx.base|, assigns integer literals the type \verb|clx.int32| while floating point literals have type \verb|clx.double|, consistent with the semantics of OpenCL and C99.
+The \verb|clx.base| object is an instance of \verb|clx.Base| that is pre-constructed for convenience.
+Alternative bases can be generated via different constructor arguments.
+For example, \verb|clx.Base(flit_t=clx.float)| is a base semantics where floating point literals have type \verb|clx.float|.
+This is useful because some OpenCL devices do not support double-precision numbers, or impose a significant performance penalty for their use.
+Indeed, to even use the \verb|double| type in OpenCL, a flag must be toggled by a \verb|#pragma| (The OpenCL target we have implemented inserts this automatically when needed, along with some other flags, and annotations like \verb|kernel|).
+Similarly, for some applications, avoiding accidental overflows is more important than performance.
+Infinite-precision integers can be implemented as an active type (not shown) and the base can be asked to use it for numeric literals.
+In all cases, this occurs by inserting a constructor call form, as described above.% the constructor call form, described above, around integers to support this.
+
+The base also initializes the context and governs the semantics of variables and assignment.
+This can also permit it to allow for the use of ``built-in'' operators that were not explicitly imported.
+The semantics of \verb|get_global_id| and \verb|printf| are provided by \verb|clx.base|.
+In fact, the base provides extended versions of them (the former permits multi-dimensional global IDs, the latter supports typechecking format strings).A base which does not provide them as built-ins (requiring that they be imported explicitly) can be generated by passing the \verb|primitives=False| flag.
+
+\subsubsection{Expressiveness}\label{examples}
+Thus far, we have focused mainly on the OpenCL target and shown examples of fairly low-level active types: those that implement OpenCL's primitives (e.g.
+\verb|clx.Ptr|) and extend them in simple but convenient ways (e.g.
+\verb|clx.Cplx|).
+
+But Ace has proven useful for more than low-level tasks like programming a GPU with OpenCL.
+We now describe some interesting extensions that implement the semantics of primitives drawn from a range of different language paradigms, to justify our claim that these mechanisms are highly expressive.
+
+\subsubsection{Growing a Statically-Typed Python Inside Ace}
+Ace comes with a target, base and type implementing Python itself: \verb|ace.python.python|, \verb|ace.python.base| and \verb|ace.python.dyn|.
+These can be supplemented by additional active types and used as the foundation for writing statically-typed Python functions.
+These functions can either be compiled ahead-of-time to an untyped Python file for execution, or be immediately executed with just-in-time compilation, just like the OpenCL examples (not shown in this proposal).% Many of the examples in the next section support this target in addition to the OpenCL target we have focused on thus far.
+
+\subsubsection{Recursive Labeled Sums}
+\begin{codelisting}
+\lstinputlisting{../ace-pldi14/datatypes_t.py}
+\caption{\texttt{[datatypes\_t.py]} An example using statically-typed functional datatypes.}
+\label{datatypest}
+\end{codelisting}
+\begin{codelisting}
+\lstinputlisting{../ace-pldi14/datatypes.py}
+\caption{\texttt{[datatypes.py]} The dynamically-typed Python code generated by running \texttt{acec datatypes\_t.py}.}
+\label{datatypes}
+\end{codelisting}
+Listing \ref{datatypest} shows an example of the use of statically-typed polymorphic recursive labeled sum types together with the Python implementation just described.
+It shows two syntactic conveniences that were not mentioned previously: 1) if no target is provided to \verb|ace.fn|, the base can provide a default target (here, \verb|ace.python.python|); 2) a concrete function can be generated immediately by providing a type assignment immediately in braces (in Python 3, a more convenient syntax is available).
+Listing \ref{datatypest} generates Listing \ref{datatypes}.
+
+Lines \ref{datatypest}.3-\ref{datatypest}.11 define a function that generates a recursive algebraic datatype representing a tree given a name and another Ace type for the data at the leaves.
+This type implemented by the active type family \verb|fp.Datatype|.
+A name for the datatype and the case names and types are provided programmatically on lines \ref{datatypest}.3-\ref{datatypest}.8.
+To support recursive datatypes, the case names are enclosed within a lambda term that will be passed a reference to the datatype itself.
+These lines also show two more active type families: units (the type containing just one value), and immutable pairs.
+Line \ref{datatypest}.13 calls this function with the Ace type for dynamic Python values, \verb|py.dyn|, to generate a type, aliased \verb|DT|.
+This type is implemented using class inheritance when the target is Python, as seen on lines \ref{datatypes}.1-\ref{datatypes}.13.
+For C-like targets, a \verb|union| type can be used (not shown).
+
+The generic function \verb|depth_gt_2| demonstrates two features.
+First, the base has been setup to treat the final expression in a top-level branch as the return value of the function, consistent with typical functional languages.
+Second, the \verb|case| ``method'' on line \ref{datatypest}.17 creatively reuses the syntax for dictionary literals to express a nested pattern matching construct.
+The patterns to the left of the colons are not treated as expressions (that is, a type and translation is never recursively assigned to them).
+Instead, the active type implements the standard algorithm for ensuring that the cases are exhaustive and not redundant \cite{pfpl}.
+If the final case were omitted, for example, the algorithm would statically indicate a warning (or optionally an error, not shown), just as in ML.
+
+%\subsubsection{Active Targets and Cross-Compilation}\label{atargets}
+%An active target, which we describe being used above, is an instance of \verb|ace.ActiveTarget|.
+It has two primary roles in Ace: 1) providing an API for code generation to a target language, providing term and type constructors for use during translation (and, optionally, a type system implementation to support representational consistency checks); 2) implementing the interoperability layer to support calling of  generic or concrete function directly from Python, as described in Sec.
+\ref{implicit}.
+%
+%An active type or base can support multiple active targets by simply examining \verb|target| during translation (e.g.
+by performing capability checks using Python's \verb|hasattr| function).
+Alternatively, the active target can simulate the interface of a different target but generate code differently.
+For example, although our implementation of OpenCL's type system as a collection of active types and an active base relies on a target that supports OpenCL terms and types, our \verb|CUDA| and \verb|C99| targets provide partial support for the same term and type forms, in order to support cross-compilation.
+We will show the latter being used in Sec.
+\ref{examples}.
+%
+
+%\subsection{Compositional Reasoning}\label{safety}
+%Every statement and expression in an Ace function is governed by exactly one active type, determined by the dispatch protocol, or by the single active base associated with the Ace function.
+The representational consistency check ensures that compilation is successful without requiring that types that abstract over other types (e.g.
+container types) know about their internal implementation details.
+Together, this permits compositional reasoning about the semantics of Ace functions even in the presence of many extensions.
+This stands on contrast to many prior approaches to extensibility, where extensions could insert themselves into the semantics in conflicting ways, making the order of imports matter (see Sec.
+\ref{related}).
+
 %Type assignment to generic functions is similar in some ways to template specialization in C++.
@@
 
-\subsection{Remaining Tasks}
+\subsubsection{Remaining Tasks and Timeline}
+A paper on this work was recently rejected from PLDI 2014.
+We plan to resubmit this work to OOPSLA 2014, due on Mar.
+25th, after completing the following tasks (largely after the Mar.
+1 ICFP deadline):
 \begin{itemize}
-\item We need to fix a minor issue raised by a reviewer in the semantics of @$\lambda$ about variables in internal terms when they are substituted.
-\item We need to write full proofs of the theorems stated in the @$\lambda$ paper.
-\item We need to develop an elaboration of a subset of the Python grammar to @$\lambda$, to connect the two.
-\item We need to detail the examples in Sec. 4 of the Ace paper more substantially.
+\item The reviewers asked for more details about the composition properties of extensions, the relationship to abstract data types, to work on staging and on how variables and contexts are treated.
+We will give more space to these issues in the new draft.
+\item We will provide more details, given the additional space available in an OOPSLA paper, of the examples other than OpenCL.
+We will consider using an example other than OpenCL to motivate the main body of the paper, so that the general distaste most PL researchers have for direct memory manipulation can be avoided.
+\item We will attempt to connect better to the theoretical work we have done on @$\lambda$, to emphasize that this work is about integrating that mechanism into an existing language (with the compromises this must necessarily entail).
+\item The reviewers made several more concrete suggestions at different points in the paper that we will integrate.
 \end{itemize}

### 2014-02-26 thesis-proposal/background.tex
More progress on Sec. 5

@@
 
-More powerful and direct compile-time \emph{term rewriting mechanisms} available in some languages can also be used for optimization, as well as for introducing specialized error checking logic and extending a language with new abstractions.
-These mechanisms are highly expressive because they allow users to programmatically manipulate syntax trees directly, but they suffer from problems of composability and safety.
+More powerful and direct compile-time \emph{term rewriting mechanisms} available in some languages can also be used for optimization, as well as for introducing specialized error checking logic and implementing new abstractions.
+These mechanisms  suffer from problems of composability, safety and expressiveness.
 For example, compile-time macros, such as those in MetaML \cite{Sheard:1999:UMS}, Template Haskell \cite{SheardPeytonJones:Haskell-02} and Scala \cite{scala-macros}, take full control over all of the code that they enclose.
 This can be problematic, however, as outer macros can interfere with the functionality of inner macros.
-Moreover, once a value escapes a macro's scope, there is no way to rely on the guarantees and features that were available within its scope, because the output of a macro is simply a value in the underlying language (a problem fundamentally related to the problem of relying on a common intermediate language,  described in Section \ref{external-approaches}).
-Thus, macros can be used to automate code generation, but not to truly extend the semantics of a language.
+Moreover, once a value escapes a macro's scope, there is no way to rely on the guarantees and features that were available within its scope, because the output of a macro is simply a term in the underlying language (a problem fundamentally related to the problem of relying on a common intermediate language,  described in Section \ref{external-approaches}).
+Thus, macros can be used to automate code generation, but not to globally extend the syntax or static guarantees of a language.
 It can also be difficult to reason about the semantics of code when any number of enclosing macros may be manipulating it, and to build tools that operate robustly in their presence.
@@
 Similarly, the derivative of Haskell implemented by the Glasgow Haskell Compiler (GHC) allows providers to introduce custom compile-time term rewriting logic if an appropriate flag is passed in \cite{jones2001playing}.
-In both cases, the user-defined logic can dispatch on arbitrary patterns of code throughout the component or program the extension is activated within, so these mechanisms are highly expressive and avoid some of the difficulties of explicitly invoked macros.
+In both cases, the user-defined logic can dispatch on arbitrary patterns of code throughout the component or program the extension is activated within, so these mechanisms permit non-local extensions to the static and dynamic semantics.
 But libraries containing such global rewriting logic cannot be safely composed because two different libraries may attempt to rewrite the same piece of code differently.
@@
 
-Another example of an active library approach to extensibility is SugarJ \cite{erdweg2011sugarj} and other languages generated by Sugar* \cite{erdweg2013framework}, like SugarHaskell \cite{erdweg2012layout}.
+Another example of an active library approach to extensibility with non-local scope is SugarJ \cite{erdweg2011sugarj} and other languages generated by Sugar* \cite{erdweg2013framework}, like SugarHaskell \cite{erdweg2012layout}.
 These languages permit libraries to extend the base syntax of the core language in a nearly arbitrary manner, and these extensions are imported transitively throughout a program.
-Unfortunately, this flexibility means that  extensions are also not safely composable.
+Unfortunately, this flexibility again means that  extensions are  not safely composable.
 For example, a library that defines a literal syntax for HTML would conflict with another that defines a literal syntax for XML because they define differing semantics for some of the same syntactic forms.

### 2014-02-26 thesis-proposal/parsing.tex
More progress on Sec. 5

@@
 \section{Type-Specific Languages}\label{aparsing}
-General-purpose abstractions are quite powerful.
-By using a general-purpose abstraction to define a data structure, one immediately benefits from a body of useful operations, established reasoning principles, well-optimized implementations and tool support.
-For example, lists can be encoded using inductive datatypes, the general-purpose abstraction mechanism that most functional programming languages emphasize.
+General-purpose abstraction mechanisms like those built into modern languages are semantically quite flexible.
+By using a general-purpose abstraction mechanism to define an abstraction, one immediately benefits from a body of useful operations, established reasoning principles, well-optimized implementations and tool support.
+For example, lists can be encoded using a more general mechanism for defining  inductive datatypes.
 Intuitively, a list can either be empty, or be broken down into a \emph{head} element and a \emph{tail}, another list.
@@
 
-Although list, number and string literals are nearly ubiquitous features of modern languages, some languages also  provide specialized notation for other common data structures (like maps and sets), data formats (like XML and JSON), query languages (like regular expressions and SQL), markup languages (like HTML) and others.
+Although list, number and string literals are nearly ubiquitous features of modern languages, some languages also  provide specialized notation for other common data structures (like maps and sets), data formats (like XML and JSON), query languages (like regular expressions and SQL), markup languages (like HTML) and many other types of data.
 For example, a language with built-in notation for HTML and SQL, with type-safe \emph{interpolation} of host language terms within curly braces, might define:
@@
 
-Unfortunately, it does not appear that this is sufficient.
+Unfortunately, it does not appear that this is working.
 Situations like this, where maintaining strong correctness, performance and security guarantees entails significant syntactic overhead, causing developers to turn to worse solutions that are more convenient, are quite common.
@@
 %%\end{lstlisting}
-This is problematic, for all of the same reasons as described above: needing explicit conversions between representations, interference issues with string syntax, correctness problems, performance overhead and security issues.
+This is problematic, for all of the reasons described above: it requires explicit conversions between representations, interference issues with string syntax, correctness problems, performance overhead and security issues.
 
@@
 In this work, we will describe an alternative parsing strategy that avoids these problems by shifting responsibility for parsing certain \emph{generic literal forms} into the typechecker.
-The typechecker, in turn, defers responsibility to named user-defined types, by treating the body of the literal as a term of the   \emph{type-specific language (TSL)} associated with the type it is being checked against.
+The typechecker, in turn, defers responsibility to user-defined types, by treating the body of the literal as a term of the   \emph{type-specific language (TSL)} associated with the type it is being checked against.
 The TSL is responsible for rewriting this term to ultimately use only general-purpose notation.
 This strategy avoids the problem of conflicting syntax, because neither the base language nor TSLs are ever extended directly.
-It also avoids semantic conflicts -- the meaning of a form like \verb|{ }| can differ depending  on its type, so it is safe to use it for empty sets, dictionaries and other data structures, like JSON literals.
-This also frees these common notations from being tied to the variant of a  data structure built into the standard library, which sometimes does not provide the exact semantics that a programmer needs (for example, Python dictionaries do not preserve key insertion order).
+It also permits semantic flexibility -- the meaning of a form like \verb|{ }| can differ depending  on its type, so it is safe to use it for empty sets, dictionaries and other data structures, like JSON literals.
+This frees these common notations from being tied to the variant of a  data structure built into a language's standard library, which sometimes does not provide the exact semantics that a programmer needs (for example, Python dictionaries do not preserve key insertion order).
 %\begin{lstlisting}
@@
 To allow us to focus on the essence of our proposal, the variant of Wyvern we will describe in this thesis is simpler than the variant previously described: it is purely functional (there are no effects other than non-termination) and it does not enforce a uniform access principle for objects (fields can be accessed directly).
-Objects are thus essentially records with simple methods (functions that are automatically given a self-reference) for convenience.
+Objects are thus a simple variant of records equipped with simple methods (functions that are automatically given a self-reference) for convenience.
 We also add recursive labeled sum types, which we call \emph{case types}, that are quite similar to datatypes in ML.
-One can refer to the version of the language described in this thesis as \emph{TSL Wyvern} when the variant is not clear.
+One can refer to the version of the language described in this thesis as \emph{TSL Wyvern}.
 
@@
 let imageBase : URL = new
-  val protocol = $\text{"}$SUShttpEUS$\texttt{"}$
-  val subdomain = $\texttt{"}$SUSimagesEUS$\texttt{"}$
+  val protocol : URLFrag = $\text{"}$SUShttpEUS$\texttt{"}$
+  val subdomain : URLFrag = $\texttt{"}$SUSimagesEUS$\texttt{"}$
   (* ... *)
@@
 The type annotation on \li{imageBase} implies that this literal's \emph{expected type} is \li{URL}, so the {body} of the literal (the characters between the angle brackets, in blue) will be governed by the \li{URL} TSL during the typechecking phase.
-This TSL will parse the body ({at compile-time}) to produce a Wyvern abstract syntax tree (AST) that explicitly instantiates a new object of type \li{URL}.
+This TSL will parse the body ({at compile-time}) to produce a Wyvern abstract syntax tree (AST) that explicitly instantiates a new object of type \li{URL} using general-purpose notation.
 
 In addition to supporting conventional notation for URLs, this TSL supports \emph{typed interpolation}
-of a Wyvern expression of type \li{URL} to form a larger URL. The interpolated term is delimited by percent signs,
+of another Wyvern expression of type \li{URL} to form a larger URL. The interpolated term is delimited by percent signs,
 as seen on line 2 of Fig.
@@
 \subsection{Layout-Delimited Literals}
+
+\begin{figure}
+\begin{lstlisting}[escapechar=$]
+casetype HTML 
+    Empty of Unit
+  | Text of String
+  | Seq of HTML * HTML 
+  | BodyElement of Attributes * HTML
+  | StyleElement  of Attributes * CSS
+  | ...
+  metadata = new : HasTSL
+    val parser : Parser = ~
+SGRM      start -> ':body'= start>
+        EGRMfn child:Exp => `SQTHTML.BodyElement(([], %EQTchildSQT%))EQT`
+SGRM      start -> ':style'= EXP>
+        EGRMfn e:Exp => `SQTHTML.StyleElement(([], %EQTeSQT%))EQT`
+SGRM      start -> '{'= EXP '}'=EGRM
+        fn e:Exp => `SQT%EQTeSQT% : HTMLEQT`
+\end{lstlisting}
+\vspace{-8px}
+\caption{A Wyvern case type with an associated TSL.}
+\label{f-htmltype}
+\end{figure}
 On line 5 of Fig.
@@
 \subsection{Implementing a TSL}
+
+\begin{figure}[t]
+\begin{subfigure}[t]{.53\textwidth}
+\begin{lstlisting}
+objtype HasTSL
+  val parser : Parser
+objtype Parser                          
+  def parse(ts : TokenStream) : (Exp * 
+    TokenStream)
+  metadata = new : HasTSL
+    val parser : Parser = (* parser generator *)      
+casetype Type
+    Var of ID
+  | Arrow of Type * Type
+  | Prod of Type * Type
+  metadata = new : HasTSL
+    val parser : Parser = (* type quasiquotes *)
+\end{lstlisting}
+\end{subfigure}
+\begin{subfigure}[t]{.55\textwidth}
+\begin{lstlisting}[linewidth=.5\textwidth]
+casetype Exp 
+    Var of ID
+  | Lam of ID * Type * Exp
+  | Ap of Exp * Exp
+  | ProdIntro of Exp * Exp
+  | CaseIntro of Type * ID * Exp
+  ... 
+  | FromTS of Tokenstream
+  | Error of ErrorMessage
+  metadata = new : HasTSL
+    val parser : Parser = (* quasiquotes *)
+\end{lstlisting}
+\end{subfigure}
+\caption{Some of the types included in the Wyvern prelude.}
+\vspace{-10px}
+\label{f-builtins}
+%\label{fig:typeParser}
+\end{figure}
 Portions of the implementation of the TSL for \li{HTML} are shown on lines 8-15 of Fig.
@@
 the CSS TSL in Fig \ref{f-example}) will operate as intended.
-
-\begin{figure}
-\begin{lstlisting}[escapechar=$]
-casetype HTML 
-    Empty of Unit
-  | Text of String
-  | Seq of HTML * HTML 
-  | BodyElement of Attributes * HTML
-  | StyleElement  of Attributes * CSS
-  | ...
-  metadata = new : HasTSL
-    val parser : Parser = ~
-SGRM      start -> ':body'= start>
-        EGRMfn child:Exp => `SQTHTML.BodyElement(([], %EQTchildSQT%))EQT`
-SGRM      start -> ':style'= EXP>
-        EGRMfn e:Exp => `SQTHTML.StyleElement(([], %EQTeSQT%))EQT`
-SGRM      start -> '{'= EXP '}'=EGRM
-        fn e:Exp => `SQT%EQTeSQT% : HTMLEQT`
-\end{lstlisting}
-\vspace{-8px}
-\caption{A Wyvern case type with an associated TSL.}
-\label{f-htmltype}
-\end{figure}
-\begin{figure}[t]
-\begin{subfigure}[t]{.53\textwidth}
-\begin{lstlisting}
-objtype HasTSL
-  val parser : Parser
-objtype Parser                          
-  def parse(ts : TokenStream) : (Exp * 
-    TokenStream)
-  metadata = new : HasTSL
-    val parser : Parser = (* parser generator *)      
-casetype Type
-    Var of ID
-  | Arrow of Type * Type
-  | Prod of Type * Type
-  metadata = new : HasTSL
-    val parser : Parser = (* type quasiquotes *)
-\end{lstlisting}
-\end{subfigure}
-\begin{subfigure}[t]{.55\textwidth}
-\begin{lstlisting}[linewidth=.5\textwidth]
-casetype Exp 
-    Var of ID
-  | Lam of ID * Type * Exp
-  | Ap of Exp * Exp
-  | ProdIntro of Exp * Exp
-  | CaseIntro of Type * ID * Exp
-  ... 
-  | FromTS of Tokenstream
-  | Error of ErrorMessage
-  metadata = new : HasTSL
-    val parser : Parser = (* quasiquotes *)
-\end{lstlisting}
-\end{subfigure}
-\caption{Some of the types included in the Wyvern prelude.}
-\vspace{-10px}
-\label{f-builtins}
-%\label{fig:typeParser}
-\end{figure}
 

### 2014-02-26 thesis-proposal/problem.tex
More progress on Sec. 5

@@
 \begin{enumerate}
-\item \textbf{Built-in syntax for pattern literals} (e.g.
-\cite{perlre}) so that the cognitive load of using regular expressions is low and malformed patterns result in intelligible {compile-time} errors.
-\item A \textbf{static and dynamic semantics} that ensures, at compile-time whenever possible, that key invariants related to regular expressions are maintained:
+\item \textbf{Syntax for pattern literals} so that the cognitive load of reading and writing patterns is low, patterns are parsed once at compile-time and malformed patterns result in intelligible {compile-time} errors.
+\item A \textbf{static semantics} that ensures that key invariants related to regular expressions are maintained:
 	\begin{enumerate}
-	\item only appropriate values are spliced into regular expressions, to avoid splicing errors and injection attacks \cite{owasp2013}
+	\item only other patterns and properly escaped strings are interpolated into a pattern, to avoid splicing errors and injection attacks \cite{owasp2013, Bravenboer:2007:PIA:1289971.1289975}
 	\item out-of-bounds backreferences are not used \cite{spishak2012type}
-	\item strings known to be in the language of a regular expression are given a type that tracks this information to ensure that string manipulation operations do not inadvertently lead to a string that is not in an assumed language \cite{fulton-thesis}.
+	\item string processing operations do not lead to a string that is malformed, when well-formedness can be captured by a regular expression \cite{fulton-thesis}
 	\end{enumerate}
-When a type error is found, an intelligible error message is provided.
+When an error is found, an intelligible error message is provided.
 %\item An \textbf{implementation} that partially or fully compiles known regular expressions into the efficient internal representation that will be used by the regular expression matching engine (e.g., a finite automata \cite{Thompson:1968:PTR:363347.363387}) ahead of time.
@@
 %In a conventional \emph{monolithic} programming system, support for each of these features would need to be built into the language and tools. 
-No system today builds in support for all of the features enumerated above.
+No system today builds in support for all of the features enumerated above in their strongest form.
 Instead, libraries generally provide support for regular expressions by leveraging  general-purpose abstraction mechanisms.
-Unfortunately, it is impossible to fully encode the syntax and the specialized static and dynamic semantics described above in terms of general-purpose notations and abstractions.
-Library providers thus need to compromise, typically by asking clients to provide regular expressions as strings and deferring parsing, typechecking and compilation to run-time.
-This introduces performance overhead and can lead to unanticipated run-time errors (as shown in \cite{spishak2012type}) and security vulnerabilities (due to injection attacks when user inputs are spliced into patterns, for example) \cite{Bravenboer:2007:PIA:1289971.1289975}.
-Useful tools for working with regular expressions are rarely integrated into editors, and even more rarely in a way that facilitates their discovery and use directly when the developer is manipulating regular expressions.
-Tools that must be discovered independently and accessed externally are used infrequently \cite{Murphy-Hill:2011:PIE:1958824.1958888} and can be more awkward than necessary \cite{Campbell:2008:DRT:1636642.1636651, Omar:2012:ACC:2337223.2337324}, leading to lower productivity.
+Unfortunately, it is impossible to fully define the syntax and the specialized static semantics described in the references above in terms of general-purpose notations and abstractions.
+Library providers thus need to compromise.
+The most common strategy is to ask clients to enter patterns as strings deferring their parsing, typechecking, compilation and use to run-time.
+This proves only a weak approximation to feature 1 (due to clashes between string escape sequences and regular expression syntax).
+None of the static guarantees can be provided in this way, leading to run-time exceptions (even in well-tested code, as shown in \cite{spishak2012type}), and logic errors that are not caught even at run-time, some of which can lead to security vulnerabilities (due to injection attacks, for example \cite{owasp2013}).
+It also introduces performance overhead (due to run-time parsing and compilation of patterns, and potentially redundant run-time string checks).
+Requiring that clients instead introduce patterns directly using general-purpose constructs like object or inductive datatype constructors or operations over an abstract type rather than via strings can only provide feature 2a while eliminating even the weak approximation to feature 1 that the use of string literals provides.
+
+None of these approaches address the issue of tool support.
+Regular expressions are not trivial to work with, so tool support can be quite helpful, even for experts, and a number of tools are available online.
+Unfortunately, tools that must be discovered independently and accessed externally are used infrequently \cite{Murphy-Hill:2011:PIE:1958824.1958888} and are less usable than editor-integrated tools \cite{Campbell:2008:DRT:1636642.1636651, Omar:2012:ACC:2337223.2337324}, leading to lower productivity.
 
@@
 \vspace{-20px}
-\caption{\small (a) When taking a language-external approach, new features are packaged together into separate languages and tools, causing problems with compatibility (described in the text).
+\caption{\small (a) When taking a language-external approach, new features are packaged together into separate languages and tools, causing problems with orthogonality and client compatibility (described in the text).
 (b) When taking a language-integrated approach, there is one extensible host language and the compile-time and edit-time logic governing new constructs is expressed within ``active'' libraries.
-It is critical that the extension mechanism  guarantees that extensions cannot interfere with one another, so that the compatibility problem is avoided.}
+If the extension mechanism guarantees that extensions cannot interfere with one another or the base system, these problems are avoided.}
 \label{approaches}
@@
 
-When the semantics or implementation of a system must be extended to fully realize a new feature, as in the example above, providers typically take a \emph{language-external approach}, either by developing a new or derivative programming system (supported by \emph{language workbenches} \cite{erdweg2013state},  \emph{DSL frameworks} \cite{fowler2010domain} or \emph{compiler generators} \cite{brooker1963compiler}), or by extending an existing system by some mechanism that is not part of the language itself, such as an extension mechanism for a {particular} compiler\footnote{Compilers that modify, or allow modification of, the semantics of their base language, rather than simply permitting semantics-preserving optimizations, should be considered a pernicious means for creating new languages.
+When the syntax and semantics of a system must be extended to fully realize a new feature, as in the example above and others we will give throughout this work, providers typically take a \emph{language-external approach}, either by developing a new or derivative programming system (supported by \emph{language workbenches} \cite{erdweg2013state},  \emph{DSL frameworks} \cite{fowler2010domain} or \emph{compiler generators} \cite{brooker1963compiler}), or by extending an existing artifact, such as an extension mechanism for a {particular} compiler\footnote{Compilers that modify, or allow modification of, the semantics of their base language, rather than simply permitting semantics-preserving optimizations, should be considered a pernicious means for creating new languages.
 Many ``practical'' compilers, including \texttt{gcc}, \texttt{GHC} and \texttt{SML/NJ}, are guilty of this, meaning that some programs that seem to be written in C, Haskell or Standard ML are actually written in tool-specific derivatives of these languages.
-Language-integrated mechanisms do not lead to such fragmentation.} or other tool.
+Language-integrated mechanisms do not lead to such fragmentation.}, editor or other tool.
 For example, a researcher interested in providing regular expression related features (let us refer to these collectively as \texttt{R}) might design a new system with built-in support for these, perhaps basing it on an existing system containing some general-purpose features (\texttt{G}).
-A different researcher developing a new language-based parallel programming abstraction or implementation strategy (\texttt{P}) might  take the same approach.
+A different researcher developing a new language-integrated parallel programming abstraction (\texttt{P}) might  take the same approach.
 A third researcher, developing an alternative parallel programming abstraction (\texttt{Q}) might again do the same.
@@
 There is no system containing \texttt{G}, \texttt{R}, \texttt{P} and \texttt{Q} in other combinations, and merging the systems containing each separately can be non-trivial because there is no requirement that providers of new features use a common mechanism.
-Even in cases where a common mechanism has been used, there are serious interference and safety issues, as we will discuss at length throughout this thesis (cf.
-Sec.
-\ref{background}, below).
+Even in cases where a common mechanism has been used, there are serious interference and safety issues, as we will discuss at length throughout this thesis.
 
-Recent evidence indicates that this is one of the major barriers preventing research from being driven into practice: developers prefer language-integrated parallel programming abstractions to library-based implementations if all else is equal \cite{cave2010comparing}, but library-based implementations are more widely adopted because ``parallel programming languages'' privilege only a few chosen  abstractions and associated implementation strategies.
-This is problematic because different parallel programming abstractions or implementation strategies are seen as more appropriate in different situations \cite{Tasharofi:2013rc}.
+Recent evidence indicates that this is one of the major barriers preventing research from being driven into practice: developers prefer language-integrated parallel programming abstractions with stronger safety guarantees and more natural syntax to library-based abstractions if all else is equal \cite{cave2010comparing}, but library-based implementations are more widely adopted because ``parallel programming languages'' privilege only a few chosen  abstractions at the language level.
+This is problematic because different parallel programming abstractions are seen as more appropriate in different situations \cite{Tasharofi:2013rc}.
 Moreover,  parallel programming support is rarely the only concern relevant to client developers outside of a classroom setting.
-Regular expression support, for example, may be simultaneously desirable for processing large amounts of textual data in parallel, but using these features together in the same compilation unit when they have been implemented by language-external means is difficult or impossible.
+Regular expression support, for example, may be simultaneously desirable for processing large amounts of textual data in parallel, but using these features together in the same compilation unit would be difficult or impossible.
 %Similarly, a language and tools designed primarily to support regular expressions might make an interesting research project, but it would not be a suitable tool for writing large applications with more varied needs.
@@
 
-\paragraph{Client Compatibility} Even in cases where for each component of a system there is a  system that is completely satisfactory, there remain problems at the interface between components.
+\paragraph{Client Compatibility} Even in cases where for each component of a software system there is a programming system that is completely satisfactory in isolation, there remain problems at the interface between components.
 An interface that exposes the specialized constructs particular to one language (e.g.
@@
 As indicated in Figure \ref{approaches}a, this only enables client compatibility in one direction.
-Calling into the common language becomes straightforward, but calling in the other direction, or between the languages sharing the common target, is not supported.
+Calling into the common language becomes straightforward and safe, but calling in the other direction, or between the languages sharing the common target, does not.
 
@@
 But many of the most innovative constructs found in modern languages (often, those that justify their creation) are difficult to define in terms of existing constructs in ways that guarantee all necessary invariants are statically maintained and that do not require large amounts boilerplate code and run-time overhead.
-For example, the type system of F\# guarantees that null values cannot occur within F\# data structures, but maintaining this important invariant still requires run-time checks because the typing rules of F\# do not apply when F\# code is called from other languages on the Common Language Infrastructure (CLI) like C\#.
-The F\# type system also includes support for checking that units of measure are used correctly \cite{syme2012expert, kennedy1994dimension}, but this highly-specialized invariant is left completely unchecked at language boundaries.
+As a trivial example with significant practical implications, F\#'s type system does not admit \verb|null| as a value for any type defined within F\# code, but maintaining this important invariant still requires run-time checks because the typing rules of F\# do not apply when F\# data structures are constructed by other languages on the Common Language Infrastructure (CLI) like C\# or SML.NET.
+This is not an issue exclusive to intermediate languages that make regrettable choices regarding \verb|null|, however.
+The F\# type system also includes support for checking that units of measure are used correctly \cite{syme2012expert, kennedy1994dimension}, but this more specialized invariant is left unchecked at language boundaries.
+No alternative  general-purpose intermediate language would change this situation.
+%Even exposing datatypes  and tuples at component boundaries, which are used ubiquitously and can be safely consumed from other languages, is not recommended when a component might be used from another language, according to drafted F\# component design guidelines \cite{fsharpguidelines}, because these are awkward to consume from other languages without the convenient primitive operations (e.g.
+pattern matching) and syntax that F\# includes.
+SML.NET prohibits exposing such types at component boundaries altogether.
 %In Scala, traits that have default method implementations are difficult to implement from Java or other JVM languages and the workaround can break if the trait is modified \cite{scalatraitinterop}. 
-In some cases, desirable features must be omitted entirely due to concerns about interoperability.
+%In some cases, desirable features must be omitted entirely due to concerns about interoperability.
 F\#, for example, aimed to retain source compatibility with Ocaml code, but due to the need for bidirectional interoperability with CLI languages, it does not support features like polymorphic variants, modules or functors \cite{ocaml-manual} because they have no apparent analogs in the type system of the CLI.
@@
 We argue that, due to these problems with orthogonality and client compatibility, taking a language-external approach to realizing a new feature should be considered harmful and avoided whenever possible.
-The goal of the research being proposed here is to design \emph{language-integrated extension mechanisms} that give providers the ability to define, within libraries, new features that have previously required central planning
+The goal of the research being proposed here is to design \emph{language-integrated extension mechanisms} that give providers the ability to define, within libraries, new features that have previously required central planning,
 %\footnote{One might compare today's programming systems to  {centrally-planned} economies, whereas extensible\- systems more closely resemble modern market economies.
@@
 
-We must proceed with caution, however: critical issues having to do with {safety} must be overcome before language-integrated extension mechanisms can be introduced into a programming system.
-If too much control over  these core aspects of the system is given  to developers, the system may become unreliable.
+We must proceed with caution, however: critical issues having to do with {safety} must be overcome before language-integrated extension mechanisms can be integrated into a system.
+If too much control over  these core aspects of the system is given  to developers, the system may become quite unreliable.
 %For example, an extension could weaken important metatheoretic guarantees previously provided by the system. 
 Type safety, for example, may not hold if the static and dynamic semantics of the language can be modified or extended arbitrarily from within libraries.
-Furthermore, even if extensions can be shown not to cause such problems in isolation, there may still be conflicts between extensions that could weaken their semantics, leading to subtle problems that only appear at link-time.
+Furthermore, even if extensions can be shown not to cause such problems in isolation, there may still be conflicts between extensions that could weaken their semantics, leading to subtle problems that only appear when two extensions are used together.
 For example, if two active libraries introduce the same syntactic form but back it with differing (but individually valid) semantics, the issue would only manifest itself when both libraries were imported somewhere within the same scope.

### 2014-02-27 att-icfp14/att-icfp14.tex
Exposition of sec 1 and 2

@@
 
+\lefthyphenmin=6
+
 \begin{document}
@@
 \begin{abstract}
-Researchers often extend an existing language with new type and operator constructors. 
-But because this is not generally possible from within the language, researchers must implement extensions as distinct forks of the language.
-This limits their utility: building applications where different components are written in languages with different type systems is unsafe and unnatural. 
-An {internally-extensible} type system could help address these issues, but designing an extension mechanism that is both expressive and reliable is a challenge: extensions must not be permitted to compromise key metatheoretic properties of the language, nor should they be allowed to interfere with one another, in any combination.
-This latter property, \emph{conservativity}, is not guaranteed by prior mechanisms.
-
-We introduce @$\lambda$, a simply-typed lambda calculus with simply-kinded type-level computation where the  semantics of new type and operator constructors can be defined in a functional style from within.
-Its semantics are structured as an extensible variant of the Harper-Stone elaboration semantics used to specify Standard ML.
-We incorporate typed compilation techniques and a form of type abstraction that together allow us to prove type safety, decidability and conservativity.
-We intend @$\lambda$ to serve as a minimal foundation for future work on extension mechanisms for typed programming languages, but it is already quite expressive.
-We begin by showing how a more natural concrete syntax can be introduced by a type-directed desugaring to @$\lambda$.
-Then, we show how a number of abstractions that must be built into other languages can be implemented orthogonally, as libraries, within @$\lambda$.
-We conclude with a discussion of its present limitations.
+Researchers often need to extend an existing language with new type and operator constructors to realize a new abstraction in its strongest form. 
+But this is not generally possible from within, so it is common to develop dialects and ``toy'' languages.
+Unfortunately, taking this approach limits the utility of these abstractions: they cannot be imported by clients of other dialects and languages, and building applications where different components rely on different type systems is both unsafe and unnatural.
+%An {internally} {extensible} type system could help address these issues, but designing an extension mechanism that is both expressive and safe is non-trivial: extensions must not be permitted to compromise key metatheoretic properties of the language, nor should they be allowed to weaken one another, in any combination.
+
+We introduce @$\lambda$, a simply-typed lambda calculus with simply-kinded type-level computation where new indexed type and operator constructors can be declared  from within.
+%It is structured in four layers: at the top-level, extension providers declare new indexed type constructors and term-level operator constructors.
+Type-level functions associated with operator constructors define their static and dynamic semantics, the latter by translation to a fixed typed internal language.
+By lifting compiler correctness techniques pioneered by the TIL compiler for Standard ML into the language, the ``actively typed semantics'' guarantees type safety.
+Going further, the semantics enforce a novel form of representation independence at extension boundaries that ensures that extensions are mutually conservative (i.e.
+they do not weaken or interfere with one another, so that they can  be used together in any combination).
+%Totality of the type-level language guarantees decidability of type checking and equality constraints on type indices makes type equivalence decidable. 
+We intend @$\lambda$ as a minimal foundation for future work on safe language-integrated extension mechanisms for typed programming languages, but it is already quite expressive.
+We demonstrate by showing how a conventional concrete syntax can be introduced by type-directed dispatch to Core @$\lambda$, then discuss a number of typed language fragments that can be introduced as orthogonal ``libraries'' (and discuss the sorts of fragments that, as yet, cannot).
 \end{abstract}
@@
 \section{Introduction}
-Typed programming languages are typically organized around a collection of indexed type and operator constructors.
-The simply-typed lambda calculus (STLC), for example, provides a single type constructor, $\rightarrow$, indexed by a pair of types, and two operator constructors, $\lambda$, indexed by a type, and $\opvar{ap}$, which can be thought of as being indexed trivially.
-We can extend it with universally quantified and recursive types by adding the type constructors $\forall$ and $\mu$, each indexed by a type binder (a type-level term that may refer to a free type variable), and their associated operator constructors.
-It is tempting to stop here: the language is now universal \todo{right terminology?} (i.e.
-Turing complete) and the dynamic semantics of operators associated with other useful types, including sums and products, can be defined by a method analagous to Church encodings in the untyped lambda calculus \cite{reynolds}.
-This is of clear theoretical and pedagogical interest.
-But this calculus is plainly impractical as a programming language.
-Reynolds, striking a cautionary tone reminiscent of Perlis in his description of the ``Turing tarpit'' \cite{perlis}, put it succinctly \cite{reynolds}:
+Typed programming languages are often described in fragments, each consisting of a small number of type constructors (often one) and associated (term-level) operator constructors.
+The simply typed lambda calculus (STLC), for example, consists of a single fragment containing a single type constructor, $\rightarrow$, indexed by a pair of types, and two operator constructors: $\lambda$, indexed by a type, and \verb|ap|, which we may think of as being indexed trivially.
+A fragment is often identified by the type constructor it is organized around, so the STLC might also be called $\mathcal{L}\{\rightarrow\}$, following the notational convention used by Harper \cite{pfpl} and others.
+
+G\"odel's $\mathbf{T}$ consists of the $\rightarrow$ fragment and the \verb|nat| fragment, defining natural numbers and a recursor that allows one to ``fold'' over a natural number.
+We might thus call it $\mathcal{L}\{\rightarrow \mathtt{nat}\}$.
+This language is  more powerful than the STLC because the STLC admits an embedding (trivially) into  $\mathbf{T}$ but the reverse is not true.
+Buoyed by this fact, we might go on by adding fragments that define sums, products and various forms of inductive types (e.g.
+lists), or perhaps a general mechanism for defining inductive types.
+Each fragment clearly increases the expressiveness of our language.
+
+If we consider the $\forall$ fragment, however, defining universal quantification over types (i.e.
+parametric polymorphism), we might take pause.
+In $\mathcal{L}\{\rightarrow\,\forall\}$, studied variously by Girard as System $\mathbf{F}$ \cite{girardF} and Reynolds as the polymorphic lambda calculus \cite{polylam}, it is known that sums, products, and inductive and co-inductive types can all be weakly defined via a technique similar to Church encodings \cite{reynolds}\todo{what to cite for this?}.
+While of substantial theoretical (and pedagogical) interest, this fact does not render irrelevant the construction of a language like $\mathcal{L}\{\rightarrow\forall~\mathtt{nat} + \times\}$.
+Reynolds, in a remark that recalls the ``Turing tarpit'' of Perlis \cite{Perl82a}, cautioned against using the existence of a weak encoding as an argument for the practicality of programming with the corresponding embedding directly \cite{Reynolds94anintroduction}:
 \begin{quote}
@@
 It is clear that the language requires a novel programming style.
+Moreover, it is likely that certain important functions cannot be expressed by their most efficient algorithms.
 \end{quote}
 
-A simple  language suffices when only the dynamic semantics of a program are being considered but continuing to add type and operator constructors (collectively, \emph{constructs}) to a language can make reasoning about programs simpler and more precise and support a more natural programming style.
-Consistent with this view, typed programming languages generally expose more constructs externally than their compilers work with internally.
-A functional language like ML, for example, might expose $n$-ary product types (tuples), labeled product types (records), recursive labeled sum types (datatypes) and forms of type abstraction (polymorphism and abstract data types).
-A compiler for the language might then translate these to an intermediate  language (IL) with only simple products and sums for further compilation to machine code.
-Correctly implementing the dynamic semantics of the language is the primary concern during this phase, so a simple IL that decreases the number of cases that must be considered when implementing and verifying the compiler is a wise choice.
-
-It is again tempting to stop here.
-Indeed, languages like ML are often referred to as ``general-purpose languages'' and have been used successfully for a range of computing tasks.
-The constructs they provide certainly occupy ``sweet spots'' in the design space, and many abstractions can be encoded using these constructs in a completely satisfactory manner.
-However, they admit no new ``universality'' theorems  stronger than those for the simple calculus described above, and situations continue to arise in both research and practice where new constructs are desirable:
-
+Finding a weak encoding of a language fragment in terms of a collection of others is useful if \emph{implementing the dynamic semantics correctly} is one's main interest, but adding new fragments directly to a language can make statically reasoning about programs more precise, support a more natural programming style and endow the language with a more favorable cost semantics.
+Consistent with this view, typed programming languages like ML expose, for example, both $n$-ary product types and record types, despite the fact that any particular product or record type is weakly definable using  polymorphic functions (or simply in terms of the other).
+The compiler for such a language, on the other hand, \emph{is} mainly concerned with implementing the dynamic semantics correctly, and indeed, compilers often use the fact that weak definability results are quite readily derivable to their advantage, translating the constructs of their external language (EL) to those of a much simpler internal language (IL) after typechecking is complete.
+%Correctly implementing the dynamic semantics of the language becomes the primary concern during this phase, so a simple IL that decreases the number of cases needing consideration when implementing and verifying the compiler is a wise choice.
+
+Having established why a minimal language like $\mathcal{L}\{\rightarrow\forall\}$, while suitable as an internal language, needs to be extended with additional fragments before it is suitable for use as a human-facing external language, we might now ask whether modern ``general-purpose'' languages like ML are fundamentally different.
+These languages have been used successfully for a wide variety of computing tasks, admitting many more strong encodings than a  calculus like $\mathcal{L}\{\rightarrow \forall\}$.
+It is certainly not our intention to argue against the notion that the particular fragments they have chosen to privilege occupy ``sweet spots'' in the design space.
+But in fact, situations continue to arise in both research and practice where the direct introduction of a new fragment is desirable because it can still only be weakly defined in terms of existing fragments, or a strong encoding is possible but impractically verbose or inefficient:
 \begin{enumerate}
-\item General-purpose constructs continue to evolve.
-There are many  variations on record types, for example: structurally-typed records, or records with forms of dynamic dispatch.
-We will show an example of records with  prototypic dispatch in Sec.
-\ref{prototypes}, and argue that they cannot be safely and naturally defined in terms of standard functional constructs.
-Sum types also admit variants, as we will discuss in the next section.
-
-\item Specialized type systems that enforce stronger invariants than general-purpose abstraction mechanisms can enforce are often proposed by researchers.
-One need not take more than a cursory glance through the literature to discover type systems for parallel programming \cite{a,b}, concurrent programming \cite{cml,a,b}, distributed programming \cite{tom7}, dataflow programming \cite{reactiveml}, authenticated data structures \cite{popl13}, information flow \cite{walker00, smith2001}, databases \cite{db}, aliasing \cite{naden12}, effects \cite{effects},  network protocols \cite{sekar99}, regular expressions \cite{regex}, units of measure \cite{keneddy} and many others.
-\item Interoperability layers that allow programmers to safely and naturally interact with libraries written in a different language require enforcing the type system of that language within the calling language.
-Using library-based interoperability layers today can lead to safety issues when this is not possible.
-%\item General-purpose abstraction mechanisms are typically implemented uniformly.
-For example, inductive datatypes are typically implemented using tagged pointers to a payload.
-In some cases (e.g.
-natural numbers) it may be desirable to implement an abstraction in an alternative manner (e.g.
-as an integer) so that certain operations perform better.
-Parallel abstractions are of particular note, as they can benefit substantially from specialized implementation strategies on specialized hardware (e.g.
-a GPU).
-Note that if control over internal implementation details were the only relevant use case, abstract datatypes might be appropriate.
+\item General-purpose abstractions continue to evolve and admit variants.
+There are many  variations on record types, for example: records with functional record update, prototypic delegation or specified field ordering (that is, labeled tuples) are either awkward to strongly define or cannot be.
+Similarly, sum types also admit variants (e.g.
+various forms of open sums).
+Even something as seemingly simple as a type-safe \verb|printf| operator requires extending a language like Ocaml  \cite{ocaml-printf}.
+
+\item Perhaps more interestingly, specialized type systems that enforce stronger invariants than general-purpose constructs are capable of enforcing are often developed by researchers.
+One need take only a brief  excursion through the literature to discover language extensions that support parallel programming \cite{a}\todo{citations}, concurrent programming \cite{cml}, distributed programming \cite{tom7}, dataflow programming \cite{reactiveml}, authenticated data structures \cite{popl13}, information flow security \cite{walker00, smith2001}, database queries \cite{db}, aliased references \cite{naden12}, network protocols \cite{sekar99}, units of measure \cite{keneddy} and a bounty of others.% All of these are implemented as dialects of existing languages, presumably because a strong encoding was not feasible.
+\item Foreign function interfaces (FFIs) that allow programmers to safely and naturally interact with libraries written in an existing language require enforcing the type system of the foreign language within the calling language.
+%Using  a FFI that does not do this can lead to safety issues, even when both languages are separately known to be safe.
+Safe FFIs generally require direct extensions to the language.
+For example, MLj extended Standard ML with constructs for safely and naturally interfacing with Java \cite{mlj}.
+For any other language, including  others on the JVM like Scala and the many languages that might be accessible via Java's native FFI, there is no way to guarantee that language-specific invariants are statically maintained, and the interface is certainly far from natural.
 \end{enumerate}
 
-Today, complete solutions to these problems generally take the form of a \emph{fork} of an existing general-purpose language (discussed further in the next section).
-This is unsatisfying because then these solutions are  \emph{non-orthogonal}: a programmer can choose either a language supporting a reasonable approach to parallel programming  or one that builds in support for working with and statically reasoning about regular expressions, but there is not necessarily a language that supports both.
-Even if a separation of such concerns can be achieved, such that each component of an application can be written using a specialized language (as advocated by proponents of the \emph{language-oriented approach} to software development \cite{language-oriented}), there will be an \emph{interoperability} problem at the interface between components, as described in item 3 above.
+These language fragments are generally disseminated as distinct \emph{dialects} of an existing general-purpose language, constructed either as a fork, a feature behind a compiler flag or, using tools like compiler generators, DSL frame\-works or  language workbenches.
+This, we argue, is quite unsatisfying: a programmer can choose either a dialect supporting an innovative approach to parallel programming or one that builds in support for statically reasoning about units of measure, but there may not be an available dialect supporting both.
+Forming such a dialect is alarmingly non-trivial, even in the rare situation where a common framework has been used or the dialects are implemented within the same compiler, as these mechanisms do not guarantee that different combinations of individually sound language fragments remain sound when combined.
+Metatheoretic and compiler correctness results can only be derived for the dialect \emph{resulting} from language   composition operations, so even  fragmentary frameworks induce an unbounded, combinatoric collection of proof obligations that providers have little choice but to  burden clients with.% We will consider this further in our discussion of related work in Sec.
+\ref{related-work}.
+
+These are not the sorts of problems usually faced by library providers.
+Well-designed languages preclude the  possibility of conflict between two libraries and ensure that the semantics of one library  cannot be weakened by another by strictly enforcing abstraction barriers.
+For example, a module declaring an abstract type in ML can rely on its representation invariants no matter which other modules are in use, so clients can assume that they will operate robustly in combination without attending to burdensome ``link-time'' proof obligations.
+
+We draw inspiration from this approach to design a mechanism that makes it possible to extend the static and dynamic semantics of an external portion of the language with new type and operator constructors from within.
+The semantics of the language as a whole constrains these extensions to maintain type safety and decidability of type checking and type equality.
+Moreover, an important class of lemmas that play a key role in  proofs that an extension enables strong encodings of a language fragment can be derived assuming a ``closed world'' (that is, in the completely traditional manner where one constructs a complete language and reasons inductively).
+The language  guarantees that they will be conserved in the  ``open world'' by enforcing  representation independence at extension boundaries.
+
+Our mechanism essentially tasks the type-level language, in a controlled manner, with implementing portions of the logic that would normally be implemented in the front-end of the compiler.
+When typechecking a user-defined operator invocation, the semantics invoke a type-level function associated with the operator's constructor to \emph{compute} both a type and a translation to a fixed typed internal language (together called a \emph{derivate}).
+Language fragments that can be weakly defined in terms of the typed internal language can be made strongly definable in the external language by introducing, in this way, the necessary  constructors and static semantics.
+We call this an \emph{actively typed semantics}.
+
+We will begin in Sec.
+\ref{atlam} by introducing a minimal calculus with an actively typed semantics, @$\lambda$.
+The external language builds in only the $\rightarrow$ fragment.
+We then use the $\mathtt{nat}$ fragment of G\"odel's $\mathbf{T}$ and a fragment defining $n$-ary products as pedagogical examples to explain how extensions are written and motivate the constraints that the semantics must impose to make the guarantees that we claim.
+We then examine the semantics in greater detail and state and sketch proofs of strong safety and conservativity theorems in Sec.
+\ref{theory}.
+\todo{organization of rest of sections}
+
+\section{@$\lambda$}
+\begin{figure*}[t]
+\small
+$$\begin{array}{rccl}	
+\textbf{programs} & \rho & ::= & \pfam{\familyDf}{\progsort} 
+%\pipe \pdef{t}{\kappa}{\tau}{\progsort} 
+\pipe e\\
+		&	\theta	&	::= &	\tops{op}{\kappaidx}{i}{a}{\taut{def}} \pipe 
+												\topp{\theta}{\theta}\\
+\\
+\textbf{external terms} 				&	e	&	::=	&	\evar{x} \pipe 
+%														\efix{x}{\tau}{e} \pipe 
+														\elam{\evar{x}}{\tau}{e} \pipe 
+														\eop{Tycon}{op}{
+															\tauidx
+														}{
+  												    		\splat{e}{1}{n}
+														} \\
+									& 		&		& 	\\
+
+\textbf{internal terms} 				& 	\iota	&	::=	&	\evar{x} \pipe 
+												\ifix{\evar{x}}{\sigma}{\iota} \pipe
+												\ilam{\evar{x}}{\sigma}{\iota} \pipe 
+												\iapp{\iota_{1}}{\iota_{2}} \pipe
+												\iintlit \pipe \iop{\iota_{1}}{\iota_{2}} \pipe \iIfEq{\iota_{1}}{\iota_{2}}{\dint}{\iota_{3}}{\iota_{4}} \\
+										& & \pipe & 
+												\iunit \pipe
+												\ipair{\iota_{1}}{\iota_{2}} \pipe 
+												\ifst{\iota} \pipe
+												\isnd{\iota} \pipe
+												 \iinl{\sigma_2}{\iota_1} \pipe \iinr{\sigma_1}{\iota_2} \pipe \icase{e}{x}{e_1}{x}{e_2} \\
+
+%\text{deabstracted}& \iota & ::= & \mathcal{G}[\iota, \sigma]\\
+\textbf{internal types}			&	\sigma	&	::=	&    \darrow{\sigma_1}{\sigma_2} \pipe \dint \pipe \dunit \pipe \dpair{\sigma_1}{\sigma_2} \pipe \dsum{\sigma_1}{\sigma_2}
+
+\\
+\\
+
+\hspace{-5pt}\textbf{type-level terms} 	& \tau 	& ::= 	& 	\tvar{t} \pipe 
+														\tlam{t}{\kappa}{\tau} \pipe 
+														\tapp{\tau_1}{\tau_2} \pipe
+														\tnil{\kappa} \pipe \tcons{\tau_1}{\tau_2} \pipe 
+									                     \tfold{\tau_1}{\tau_2}{h}{t}{r}{\tau_3}
+														\\
+
+	 			& 		& \pipe	& 	 \iintlit \pipe \iop{\tau_1}{\tau_2} \pipe \tlabel{label} \pipe \tunit \pipe 
+														\tpair{\tau_{1}}{\tau_{2}} \pipe 
+														\tfst{\tau} \pipe 
+														\tsnd{\tau} 
+														\\	
+									&       & \pipe & \tinl{\kappa_2}{\tau_1} \pipe \tinr{\kappa_1}{\tau_2} \pipe \tsumcase{\tau}{x}{\tau_1}{x}{\tau_2} \\
+\text{equality}  & & \pipe & 					\tifeq{\tau_{1}}{\tau_{2}}{\kappa}{\tau_{3}}{\tau_{4}} 
+														\\													\text{types} 						& 		& \pipe	& 	\ttypestd \pipe \tfamcase{\tau}{Tycon}{x}{\tau_1}{\tau_2}\\
+						%				& & \pipe & \tfamcase{\tau}{Fam}{x}{\tau_1}{\tau_2}\\
+
+\text{derivates} 				& 		 & 	\pipe	&	\tden{\tauiterm}{\tautype}  \pipe \tderbind{x}{t}{\tau}{\tau_1}\\% \tdencase{\tau}{x}{t}{\tau_1}{\tau_2}\\
+ %& & \pipe & 
+%														\tdencase{\tau}{y}{x}{\tau_1}{\tau_2}
+%														 \\
+
+\text{reified IL}		&		&	\pipe	&	\titerm{\bar \iota} \pipe \titype{\bar \sigma} \\
+% &  & \pipe & \tvalof{\tau_1}{\tau_2} \pipe \iup{\tau} \\
+% 												\trepof{\tau} \pipe \dup{\tau}\\
+	& \bar{\iota} & ::= & x \pipe \ifix{x}{\bar \sigma}{\bar \iota} 
+	%\pipe \ilam{x}{\bar \sigma}{\bar \iota} \pipe \iapp{\bar \iota_1}{\bar \iota_2} 
+	\pipe \cdots \pipe \iup{\tau} \pipe \tvalof{\tauiterm}{\tautype} \\						
+ & \bar{\sigma} & ::= & \darrow{\bar \sigma_1}{\bar \sigma_2} \pipe \cdots \pipe \dup{\tau} \pipe \trepof{\tau} \\
+											\\
+\textbf{kinds} 					& \kappa	&	::=	&	\karrow{\kappa_1}{\kappa_2} \pipe \klist{\kappa} \pipe \kint \pipe
+											    \klabel \pipe
+												\kunit \pipe 
+												\kpair{\kappa_{1}}{\kappa_{2}} \pipe 
+												\ksum{\kappa_1}{\kappa_2} \pipe
+												\kTypeBlur \pipe \kDen \pipe 
+												\kITerm \pipe \kIType
+												\\
+\\												
+%\textbf{ops signature}			& \Theta	&	::=	&	\kOpEmpty \pipe \kOp{\Theta}{op}{\kappai}\\
+%											 							&		&		&	\\
+\end{array}$$
+\vspace{-10pt}
+\caption{\small Syntax of Core \atlam.
+Here, $x$ ranges over external and internal language variables, $\tvar{t}$ ranges over type-level variables, $\fvar{Tycon}$ ranges over type constructor names, $\opvar{op}$ ranges over operator constructor names, $\iintlit$ ranges over integer literals, $\tlabel{label}$ ranges over label literals (see text) and $\oplus$ ranges over the standard total binary operations over integers (e.g.
+addition).
+The form $\tvalof{\tauiterm}{\tautype}$ is  used internally by the semantics (it need not be supported by a parser).%The productions related to the internal language are written using generators $\mathcal{G}$ and $\mathcal{S}$ to avoid duplicating the syntax of common terms.
+\vspace{-10pt}
+\label{grammar}}
+\end{figure*}
+\begin{figure}[t]
+\small
+\begin{flalign}
+\label{natfam}&\family{Nat}{\kunit}{\\
+\label{z}&\quad\tops{z}{\kunit}{i}{a}{\tlam{idx}{\kunit}{\tlam{args}{\klist{\kDen}}{
+	\tapp{\tapp{\tvar{if\_empty}}{\tvar{args}}}{
+		\tden{\titerm{0}}{\ttype{Nat}{\tunit}}
+	}}}};\\
+\label{s}&\quad\tops{s}{\kunit}{i}{a}{\tlam{idx}{\kunit}{\tlam{args}{\klist{\kDen}}{\\
+& \quad\quad
+	\tapp{
+		\tapp{\tvar{pop\_final}}{\tvar{args}}}{\tlam{x}{\kITerm}{\tlam{t}{\kTypeBlur}{
+			\\
+			&\quad\quad\tapp{\tapp{\tapp{\tvar{check\_type}}{\tvar{t}}}{\ttype{Nat}{\tunit}}}{
+				\tden{\titerm{\iup{\tvar{x}}+1}}{\ttype{Nat}{\tunit}}
+			}
+		}}
+		}
+	}}};\\
+\label{rec}&\quad\tops{rec}{\kunit}{i}{args}{\tlam{idx}{\kunit}{\tlam{args}{\klist{\kDen}}{\\
+& \quad\quad
+	\tapp{\tapp{\tvar{pop}}{\tvar{args}}}{\tlam{x1}{\kITerm}{\tlam{t1}{\kTypeBlur}{\tlam{args}{\klist{\kDen}}{\\
+	&\quad\quad\tapp{\tapp{\tvar{pop}}{\tvar{args}}}{\tlam{x2}{\kITerm}{\tlam{t2}{\kTypeBlur}{\tlam{args}{\klist{\kDen}}{\\
+	&\quad\quad \tapp{\tapp{\tvar{pop\_final}}{\tvar{args}}}{\tlam{x3}{\kITerm}{\tlam{t3}{\kTypeBlur}{\\
+	&\quad\quad \tapp{\tapp{\tapp{\tvar{check\_type}}{\tvar{t1}}}{\ttype{Nat}{\tunit}}}{(\\
+	&\quad\quad \tapp{\tapp{\tapp{\tvar{check\_type}}{\tvar{t3}}}{\ttype{Arrow}{(\ttype{Nat}{\tunit},\ttype{Arrow}{(\tvar{t2},\tvar{t2})})}}}{\\
+		\label{fix}&\quad\quad \tden{\titerm{\iapp{(\ifix{f}{\darrow{\dint}{\trepof{\tvar{t2}}}}{\ilam{x}{\dint}{\\
+		\label{lastop}&\quad\quad\quad \iIfEq{x}{0}{\dint}{\iup{\tvar{x2}}}{\iapp{\iapp{\iup{\tvar{x3}}}{(x-1)}}{(\iapp{f}{(x-1)})}}}})}{\iup{\tvar{x1}}}}}{\tvar{t2}}
+	)}}
+	}}}}}}}}}}}}}}
+\\
+&}{XXX}{i}{\tlam{idx}{\kunit}{\titype{\dint}}};\\
+\label{nattype}&\etdef{nat}{\kTypeBlur}{\ttype{Nat}{\tunit}}{\\
+& \elet{plus}{\ttype{Arrow}{(\tvar{nat}, \ttype{Arrow}{(\tvar{nat}, \tvar{nat})})}}{
+	\elam{x}{\tvar{nat}}{
+		\elam{y}{\tvar{nat}}{
+			\\&\quad \eop{Nat}{rec}{\tunit}{x; y; \elam{p}{\tvar{nat}}{\elam{r}{\tvar{nat}}{
+				\eop{Nat}{s}{\tunit}{r}
+			}}}
+		}
+	}
+}{\\
+& \elet{two}{\tvar{nat}}{\eop{Nat}{s}{\tunit}{\eop{Nat}{s}{\tunit}{\eop{Nat}{z}{\tunit}{}}}}{\\
+& \eop{Arrow}{ap}{\tunit}{plus; \eop{Arrow}{ap}{\tunit}{two; two}}
+}}}
+%{\\&\elet{two}{\tvar{nat}}{\eop{Nat}{s}{\tunit}{\eop{Nat}{s}{\tunit}{\eop{Nat}{z}{\tunit}{ }}}}{\eapp{\eapp{plus}{two}}{two}}}
+%}
+%\elam{x}{\tvar{nat}}{\elam{y}{\tvar{nat}}{\\
+	%&\quad \eop{Nat}{rec}{\tunit}{x; y; \elam{p}{\tvar{nat}}{\elam{r}{\tvar{nat}}{
+	%\eop{Nat}{s}{\tunit}{r}
+	%}}}}
+\end{flalign}
+\vspace{-15px}
+\caption{\small G\"odel's $\mathbf{T}$ in \atlam, used to calculate 2+2.
+The simple helper functions $\tvar{if\_empty}$, $\tvar{pop}$, $\tvar{pop\_final}$, $\tvar{check\_type}$ all have return kind $\ksum{\kDen}{\kunit}$ (see text).
+Their definitions can be found in the appendix.
+We use \textsf{let} to bind both external and type-level variables for clarity of presentation; these can be eliminated by manually performing the indicated substitutions (or added to the semantics).}
+\label{nat}
+\end{figure}
+%
+%\begin{figure}[t]
+%\begin{lstlisting}
+%tycon Nat of 1 with  
+%    schema $\lambda$idx:1.$\blacktriangledown$($\mathbb{Z}$) 
+%    opcon Z of 1 ($\lambda$idx:1.$\lambda$args:list[Elab].is_empty args $\llbracket$$\triangledown$(0) as Nat[()]$\rrbracket$)
+%    opcon S of 1 ($\lambda$idx:1.$\lambda$args:list[Elab].pop_final args $\lambda$x:ITm.$\lambda$ty:$\star$. 
+%      checktype ty Nat[()] $\llbracket$$\triangledown$($\vartriangle$(x)+1) as Nat[()]$\rrbracket$)
+%    opcon Rec of 1 ($\lambda$idx:1.$\lambda$args:list[Elab].
+%      pop args $\lambda$x1:ITm.$\lambda$ty1:$\star$.$\lambda$args':list[Elab].
+%      pop args' $\lambda$x2:ITm.$\lambda$ty2:$\star$.$\lambda$args'':list[Elab].
+%      pop_final args'' $\lambda$x3:ITm.$\lambda$ty3:$\star$.
+%      check_type t1 Nat[()] (
+%      check_type t3 Arrow[(Nat[()], Arrow[(t2, t2)])] 
+%      	$\llbracket$$\triangledown$((fix f:$\mathbb{Z} \rightarrow$ rep(t2) is $\lambda$x:$\mathbb{Z}$.
+%	        if x = 0 then $\vartriangle$(x2) else $\vartriangle$(x3) (x - 1) (f (x - 1))) $\vartriangle$(x1)) as t2$\rrbracket$))
+%end
+%\end{lstlisting}
+%\caption{An implementation of primitive natural numbers as internal integers in @$\lambda$.}
+%\label{nat-atlam}
+%\end{figure}
+In this section, we will develop an ``actively typed'' version of the simply-typed lambda calculus with simply-kinded type-level computation called @$\lambda$.
+More specifically, the level of types, $\tau$, will itself form a simply-typed lambda calculus.
+\emph{Kinds} classify type-level terms in the same way that types conventionally classify expressions.
+Types become just one kind  of type-level value (which we will write $\kTypeBlur$, though it is also variously written $\star$, \verb|T| and \verb|Type| in various settings).
+Rather than there being a fixed set of type constructors, we allow the programmer to declare new type  constructors, and give the static and dynamic semantics of their associated operators, by writing type-level functions.
+In the semantics for this calculus, our kind system combined with techniques borrowed from the typed compilation literature and a form of type abstraction allow us to prove strong type safety, decidability and conservativity theorems.
+
+The syntax of @$\lambda$ is given in Fig.
+\ref{grammar}.
+An example of a program containing an extension that implements the static and dynamic semantics of G\"odel's \textbf{T} is given in Fig.
+\ref{nat}.
+We do not, of course, deny that natural numbers can be strongly encoded with a similar usage and asymptotic performance profile (but with potentially relevant additional function call overhead) by existing means (e.g.
+by an abstract type).
+We will provide more sophisticated examples where this is not the case below.
+%It consists of:
+%a type constructor declaration, $\fvar{Nat}$, indexed trivially, together with three operator constructors, also all indexed trivially, that implement the standard introductory forms for natural numbers as well as the recursor operator (as in G\"odel's T \cite{pfpl}).
+Following the type constructor declaration, we apply $\fvar{Nat}$ with the trivial index, $\tunit$, to form the type $\tvar{nat}$.
+Finally, we write an external term that uses the operators associated with $\fvar{nat}$ and the built-in constructor $\fvar{Parr}$, governing partial functions, to define an addition function and compute the addition of the natural numbers  two and two.
+We will introduce a more convenient concrete syntax in later portions of this thesis; for now we will restrict ourselves to the abstract syntax so that this example can directly aid in understanding the semantics.
+
+\subsection{Programs}
+
+A program, $\rho$, consists of a series of static declarations followed by an external term, $e$.
+The syntax for external terms contains three core forms: variables, functions, and a form for all other operator invocations, which we will discuss below.
+Compiling a program consists of first \emph{kind checking} its static declarations and type-level terms then \emph{type checking} the external term and, simultaneously, \emph{translating} it to a term, $\iota$, in a \emph{typed internal language}.
+In our calculus, the internal language contains partial functions (via the generic fixpoint operator of Plotkin's PCF \cite{PCF}), simple product and sum types and a base type of integers (to make our example interesting and give a nod to practicality on contemporary machines).
+In practice, the internal language could be any intermediate language for which type safety and decidability of typechecking are known, as @$\lambda$ requires these to establish the corresponding theorems about the language as a whole.
+
+The \emph{active typing judgement} relates an external term to a \emph{type} and an internal term, called its \emph{translation}, under \emph{typing context} $\Gamma$ and \emph{constructor context} $\Phi$:
+$$\ecompilesX{e}{\tau}{\iota}$$
+
+This judgement follows standard conventions for the typing judgement of a simply-typed lambda calculus in most respects.
+The typing context $\Gamma$ maps variables to types and obeys standard structural properties.
+The key novelties here are that the available type and operator constructors are not fixed by the language, but rather are tracked by the constructor context, $\fvalCtx$, and that a translation is simultaneously derived.
+The dynamic semantics of external terms are defined by their translation to the internal language.
+For this judgement, the two contexts and the external term can be thought of as input, while the type and translation are output (we do not consider a bidirectional approach in this work, though this may be a promising avenue for future research).
+
+\subsection{Types}
+User-defined type constructors are declared at the top-level of a program (or package, in a practical implementation) using \textsf{tycon}.
+Each type constructor in the program must have a unique name, written e.g.
+\fvar{Nat}.
+We do not intrinsically consider the issue of namespacing, under the assumption that globally unique names can be generated by some extrinsic mechanism (e.g.
+a URI-based scheme).
+A type constructor must also declare an \emph{index kind}, $\kappaidx$.
+Types themselves are type-level values of kind $\kTypeBlur$ and are introduced by applying a previously declared type constructor to a type-level term of its index kind, $\ttype{Tycon}{\tauidx}$.
+The kind $\kTypeBlur$ also has an elimination form: a type can be case analyzed against a type constructor in scope to extract its index.
+Like open datatypes, there is no longer a notion of exhaustiveness.
+To maintain totality, we require the default case.
+
+To permit the implementation of interesting type systems, the type-level language includes several other kinds of data alongside types.
+We lift several  standard functional constructs to the type level: unit ($\kunit$), binary sums ($\ksum{\kappa_1}{\kappa_2}$), binary products ($\kpair{\kappa_1}{\kappa_2}$), lists ($\klist{\kappa}$) and integers ($\dint$).
+We also include labels ($\klabel$), written in a slanted font, e.g.
+$\tlabel{myLabel}$, which are atomic string-like values that can only be compared, here only for equality, and play a distinguished role in the expanded syntax, as we will later discuss.
+Our first example, $\fvar{Nat}$, is indexed trivially, i.e.
+by unit kind, $\kunit$, because there is only one natural number type, $\ttype{Nat}{\tunit}$, but we will show examples of type constructors that are indexed in more interesting ways in later portions of this work.
+For example, $\fvar{Tuple}$ is indexed by a list of types and $\fvar{LabeledTuple}$ is indexed by a list pairing labels with types.
+
+Two type-level terms of kind $\kTypeBlur$ are equivalent if they apply the same constructor, identified by name, to equivalent indices.
+Going further, we ensure that deciding type equivalence requires only checking for syntactic equality after evaluation to normal form by imposing the restriction that type constructors can only be indexed by kinds for which equivalence can be decided in this way.
+All combinations of kinds introduced thus far  have this property and this is sufficient for many interesting examples.
+We leave  introducing a richer notion of  equivalence of types that  extension providers can extend as future work, as the metatheoretic guarantee that typing respects type equivalence would be quite a bit more complex in such a setting.% (a na\"ive approach to this would impose non-trivial extrinsic proof obligations onto extension developers that, unlike in others in this thesis, could threaten type safety).
+
+Type constructors are not first-class; they do not themselves have arrow kind as in some kind systems (e.g.
+\cite{Watkins08}\todo{cite}; Ch.
+22 of \emph{PFPL} describes a related system \cite{pfpl}).
+The type-level language does, however, include total functions of conventional arrow kind, $\karrow{\kappa_1}{\kappa_2}$.
+Type constructor application can be wrapped in a type-level function to emulate first-class type constructors (and indeed, such a wrapper could be generated automatically, though we avoid this for simplicity in our semantics).
+Equivalence at arrow kind does not coincide with $\beta$-equivalence, so type-level functions cannot appear in type indices.
+Our treatment of equivalence in the type-level language is thus quite similar to the treatment of term-level equality using ``equality types'' in a language like Standard ML.
+Indeed, one might imagine that a practical implementation of our calculus would start with a pure, total subset of the term language of ML to construct the type-level language (consistent with its early development as a ``metalanguage'' and, as we will see, the role of our type-level language as a DSL for implementing type systems and translators, which functional languages are widely considered well-suited for as-is).
+We leave the development of such an implementation (and of bootstrapping type-level and internal languages that are themselves user-defined or extensible) as areas for future work.
+\subsection{Operators}
+User-defined operator constructors are declared using \textsf{opcon}.
+For reasons that we will discuss, our calculus associates every operator constructor with a type constructor.
+The \emph{fully-qualified name} of the operator constructor, e.g.
+$\fvar{Nat}.\opvar{z}$, must be unique.
+Operator constructors are indexed by type-level values, like type constructors, and so also declare an index kind $\kappaidx$.
+In our first example, all the operator constructors are indexed trivially, but later examples will use more interesting indices.
+For example, the projection operators for tuples and labeled tuples use numbers and labels as indices, respectively.
+Note that neither operator constructors nor operators are first-class type-level values (we plan to investigate  lifting the latter restriction) and we do not impose any restrictions on their index kinds.
+
+In the external language, an operator is selected and invoked by applying an operator constructor to a type-level index and $n \geq 0$ \emph{arguments}, $\eop{Tycon}{op}{\tauidx}{\splat{e}{1}{n}}$.
+For example, on line 18 of Fig.
+\ref{nat}, we see the operator constructors $\fvar{Nat}.\opvar{z}$ and $\fvar{Nat}.\opvar{s}$ being invoked to compute two.\footnote{Although our focus here is entirely on semantics, a brief note on syntax: in the expanded syntax, the trivial indices and empty argument lists can be omitted, so we could write \texttt{Nat.s(Nat.s(Nat.z))}.
+With the ability to ``open'' a type's operators into the context, we could shorten this still to \texttt{s(s(z))}.
+Alternatively, with the ability to define a TSL in a manner similar to that in Sec.
+\ref{aparsing}, we might instead just write \texttt{2}.}
+To derive the active typing judgement for this form, the semantics invokes  
+ the operator constructor's  \emph{definition}: a type-level function that must examine the provided operator index and the recursively determined \emph{derivates} of the arguments to determine a derivate for the operation as a whole,  or indicate an error.
+A derivate is a type-level representation of the result of deriving the active typing judgement\footnote{Technically, the abstract active typing judgement, which is similar in form and we will introduce shortly.}: an internal term, called the {translation},  paired with a type.
+Derivates have kind $\kDen$ and introductory form $\tden{\tautrans}{\tautype}$.
+The elimination form $\tderbind{x}{t}{\tau}{\tau'}$ extracts the translation and type from the derivate $\tau$, binding it to $\tvar{x}$ and $\tvar{t}$ respectively in $\tau'$.
+%Indeed, one might be tempted to simply define derivates as products.
+However, we treat translations with much care throughout the calculus, so this is not quite the case (we will return to this shortly).
+Because constructing a derivate is not always possible (e.g.
+when there is a type error in the client's code, or an invalid index was provided), the return kind of the function is an ``option kind'', $\ksum{\kDen}{\kunit}$, where the trivial case indicates a statically-detected error in the client's program.
+In practice, it would instead require providers to report information about the precise location of the error and an appropriate error message.
+
+A translation, as we have said, is an internal term.
+To construct an internal term using a type-level function, as we must do to construct a derivate, we must expose a type-level representation of the internal language.
+A \emph{quoted internal term} is a type-level value of kind $\kITerm$ with introductory form $\titerm{\bar \iota}$ and a \emph{quoted internal type} is a type-level value of kind $\kIType$ with introductory form $\titype{\bar \sigma}$.
+Neither kind has an elimination form.
+Instead, the syntax for the quoted  internal language includes complementary \emph{unquote forms} $\iup{\tau}$ and $\dup{\tau}$ that permit the interpolation of another quoted term or type, respectively, into the one being formed.
+Interpolation is capture-avoiding, as our semantics will clarify.
+We have now described all the kinds in our language.
+
+The definition of $\fvar{Nat}.\opvar{z}$ is quite simple: it returns the derivate $\tden{\titerm{0}}{\ttype{Nat}{\tunit}}$ if no arguments were provided, and indicates an error (an \emph{arity error}, though we do not distinguish this in our calculus) otherwise.
+This is done by calling a simple helper function, $\tvar{is\_empty} : \karrow{\klist{\kDen}}{\karrow{\kDen}{(\ksum{\kDen}{\kunit})}}$, that selects the appropriate case of the sum given the derivate that should be produced if the list is indeed empty.
+The definition of $\fvar{Nat}.\opvar{s}$ is only slightly more complex, because it requires inspecting a single argument.
+The helper function $\tvar{pop\_final} : \karrow{\karrow{\klist{\kDen}}{(\karrow{\kTypeBlur}{\karrow{\kITerm}{(\ksum{\kDen}{\kunit})})}}}{(\ksum{\kDen}{\kunit})}$ ``pops'' a derivate from the head of the list and, if no other arguments remain, passes its translation and type to the ``continuation'', returning the error case otherwise.
+The continuation checks if the argument type is equal to $\ttype{Nat}{\tunit}$ using $\tvar{check\_type} : \karrow{\kTypeBlur}{\karrow{\kTypeBlur}{\karrow{\kDen}{(\ksum{\kDen}{\kunit})}}}$, which operates similarly.
+If these arity and type checks succeed, the resulting derivate is composed by adding one to the translation of the argument, passed into the continuation as $\tvar{x}$ in our example, and pairing it with the natural number type: $\tden{\titerm{\iup{\tvar{x}} + 1}}{\ttype{Nat}{\tunit}}$.
+We will return to the definition of $\fvar{Nat}.\opvar{rec}$ after in the next subsection.
+
+By writing the typechecking and translation logic in this way, as a total function, we are taking a rather ``implementation-focused view'' rather than attempting to extract such a function from a declarative specification.
+That is,  we leave to the provider (or to a program generator or kind-specific language that transforms such a specification into an operator definition) the problem of finding a deterministic algorithm that adequately implements their intended semantics, allowing us to prove decidability of typechecking for the language as a whole by essentially just citing the termination theorem for the type-level language.
+We also avoid difficulties with error reporting in practice.
+Rob Simmons' undergraduate thesis has a good discussion of these issues \cite{rjs-princeton}.
+
+Because the input to an operator definition is the recursively determined derivate of the argument in the same context as the operator appears in, our mechanism does not presently permit the definition of operator constructors that bind variables themselves or require a different form of typing judgement (e.g.
+additional forms of contexts).
+This is also why the $\lambda$ operator constructor needs to be built in.
+It is the only operator in our language that can be used to bind variables.
+The type constructor $\fvar{Arrow}$ can be defined from within the language (and is included in the ``prelude'' constructor context), but only defines the operator constructor, $\opvar{ap}$.
+These two operators translate to their corresponding forms in the internal language directly, as we will discuss below.
+We leave adding the ability to declare and manipulate new contexts as future work.
+
+\subsection{Representational Consistency Implies Type Safety}
+In our example, natural numbers are represented internally as integers.
+Were this not the case -- if, for example, we added an operator constructor $\fvar{Nat}.\opvar{z2}$ that produced the derivate $\tden{\titerm{\iinl{\dint}{\iunit}}}{\ttype{Nat}{\tunit}}$, then there would be two different internal types, $\dint$ and $\dsum{\dunit}{\dint}$,  associated with a single external type, $\ttype{Nat}{\tunit}$.
+This makes it impossible to operate compositionally on the translation of an external term of type $\ttype{Nat}{\tunit}$, so our implementation of $\fvar{Nat}.\opvar{s}$ would produce ill-typed translations in some cases but not others.
+Similarly, we wouldn't be able to write functions over all natural numbers because there would not be a well-typed translation to give to such a function.
+
+To reason compositionally about the semantics of well-typed external terms when they are given meaning by translation to a typed internal language, we must have the following property: for every  type, $\tau$, there must exist an internal type, $\sigma$, called its \emph{representation type}, such that the translation of every external term of type $\tau$ has internal type $\sigma$.
+This principle of \emph{representational consistency} arises essentially as a strengthening of the inductive hypothesis necessary to prove that all well-typed external terms translate to well-typed internal terms, precisely because operators like $\lambda$ and $\fvar{Nat}.\opvar{s}$ are defined compositionally.
+It is closely related to the concept of \emph{type-preserving compilation} developed by Morrisett et al.
+for the TIL compiler for Standard ML \cite{TIL}\todo{cite}.
+%That is, instead of being used as a necessary condition for compiler correctness, we are using it as a sufficient condition for type safety.
+%We emphasize the distinction between translation (which in our calculus endows terms with a dynamic semantics) and compilation (which must preserve the dynamic semantics of terms).
+%In the presence of extensions, checking for representational consistency becomes subtle if we wish to guarantee conservativity, as we will discuss in the next subsection.
+
+It is easy to show by induction that, under a ``closed-world assumption'' where the only available operators are $\fvar{Nat}.\opvar{z}$ and $\fvar{Nat}.\opvar{s}$, the representation type of $\ttype{Nat}{\tunit}$ is $\dint$.
+If we can maintain this under an ``open-world assumption'' (requiring that, for example, applications of operator constructors like $\fvar{Nat}.\opvar{z2}$, above, are not well-typed), and we target a type safe internal language, then we will achieve type safety: well-typed external terms cannot go wrong, because they always translate to well-typed internal terms, which cannot go wrong.
+For the semantics to ensure that representational consistency is maintained by all operator definitions, we require that each type constructor must declare a \emph{representation schema} with the keyword \textsf{schema}.
+This must be a type-level function of kind $\karrow{\kappaidx}{\kIType}$, where $\kappaidx$ is the index kind of the type constructor.
+When the compiler needs to determine the representation type of the type $\ttype{Tycon}{\tauidx}$ it simply applies the representation schema of $\fvar{Tycon}$ to ${\tauidx}$.
+
+As described above, the kind $\kIType$ has introductory form $\titype{\bar \sigma}$ and no elimination form.
+There are two forms in $\bar \sigma$ that do not correspond to forms in $\sigma$ that allow quoted internal types to be formed compositionally:
+\begin{enumerate}
+\item $\dup{\tau}$, already described, unquotes (or ``interpolates'') the quoted internal type $\tau$
+\item $\trepof{\tau}$ refers to the representation type of type $\tau$ (we will see in the next subsection why this needs to be in the syntax for $\bar \sigma$ and not directly in the type-level language)
+\end{enumerate}
+
+These additional forms are not needed by the representation schema of $\fvar{Nat}$  because it is trivially indexed.
+In Fig.
+\ref{tuple}, we show an example of the type constructor $\fvar{Tuple}$, implementing the semantics of $n$-tuples by translation to nested binary products.
+Here, the representation schema requires referring to the representations of the tuple's constituent types, given in the type index\todo{add this example from appendix of ESOP}.
+
+Operator constructor definitions might also need to refer to the representation of a type.
+We see this in the definition of the recursor on natural numbers, $\fvar{Nat}.\opvar{rec}$.
+After checking the arity and extracting the types and translations of its three arguments, it produces a derivate that implements the necessarily recursive dynamic semantics using a fixpoint computation in the internal language.
+This fixpoint computation produces a result of some arbitrary type, $\tvar{t2}$.
+We cannot know what the representation type of $\tvar{t2}$ is, we refer to it abstractly using $\trepof{\tvar{t2}}$.
+The translation is well-typed no matter what the representation type is.
+In other words, a proof of representational consistency of the derivate produced by the definition of $\fvar{Nat}.\opvar{rec}$ is parametric (in the metamathematics) over the representation type of $\tvar{t2}$.
+This leads us into the concept of conservativity.
+
+\subsection{Representational Independence Implies Conservativity}
+In isolation, our definition of natural numbers can be shown to be a strong encoding of the semantics of G\"odel's $\mathbf{T}$.
+That is, there is a bijection between the terms and types of the two languages and the typing judgements preserve this mapping.
+Moreover, we can show by a bisimulation argument that the translation produced by @$\lambda$ implements the dynamic semantics of G\"odel's $\mathbf{T}$.
+We will provide more details on this later.
+A necessary lemma in the proof, however, is that the value of every translation of an external term of type $\ttype{Nat}{\tunit}$ is a non-negative integer.
+This is needed to show that the fixpoint computation in the definition of $\fvar{Nat}.\opvar{rec}$ is terminating, as the recursor always does in $\mathbf{T}$.
+A very similar lemma is needed to show that the translation of every external term of type $\ttype{Nat}{\tunit}$ is equivalent to the translation that would be produced by some combination of applications of $\fvar{Nat}.\opvar{z}$ and $\fvar{Nat}.\opvar{s}$ (the analog to a canonical forms lemma in our setting).
+Fortunately, the definitions we have provided thus far admit such lemmas.
+
+However, these theorems are all quite precarious because they rely on exhaustive induction over the available operators.
+As soon as we load another library into the program, these theorems may no longer be \emph{conserved}.
+For example, the following operator declaration in some other type that we have loaded would topple our house of cards\todo{TBC}:
+
+%At best, they can aim to achieve a separation of concerns, such that each component of an application can be written using a specialized language (as advocated by proponents of the \emph{language-oriented approach} to software development \cite{language-oriented}), but there will still be an {interoperability} problem at the interface between components, as described above, unless an FFI has been explicitly developed for each pair of languages that might interface with each other, or library interfaces are constrained to only use a small subset of common constructs (as suggested in draft guidelines for F\#, for example \cite{fsharp-guidelines}) reducing  expressiveness and safety.
 
@@
 %\end{figure}
-
-\todo{everything below is todo}Rather than specifying a complete language, a researcher might simply specify the new type and operator constructors and
 
@@
 
-An extensible programming language could address these problems by providing a language-integrated mechanism for introducing new type and operator constructors and implementing their associated static and dynamic semantics directly.
-%Library developers need only consider which abstractions are most appropriate for their domain, without also considering whether these constructs can be exposed using abstractions appropriate to the domains of client code.
+%An extensible programming language could address these problems by providing a language-integrated mechanism for introducing new type and operator constructors and implementing their associated static and dynamic semantics directly.
+%%Library developers need only consider which abstractions are most appropriate for their domain, without also considering whether these constructs can be exposed using abstractions appropriate to the domains of client code.
 Clients can simply import any necessary constructs when using a library that relies on them, preserving safety and ease-of-use without the use of  wrappers and glue code.
 We show this competing approach in Figure \ref{approaches}(b).
-%Researchers and domain experts thus gain the ability to distribute new ideas for evaluation to a broader development community without requiring the approval of maintainers of mainstream languages, large-scale porting of code or explicit interoperability layers.
-But, as mentioned in Section \ref{language-integrated-approaches}, some significant challenges must be addressed before such a mechanism can be relied upon.
+%%Researchers and domain experts thus gain the ability to distribute new ideas for evaluation to a broader development community without requiring the approval of maintainers of mainstream languages, large-scale porting of code or explicit interoperability layers.
+%But, as mentioned in Section \ref{language-integrated-approaches}, some significant challenges must be addressed before such a mechanism can be relied upon.
 The desire for expressiveness must be balanced against  concerns about maintaining various safety properties in the presence of arbitrary combinations of user-defined  extensions to the language's core semantics.
@@
 Wyvern \cite{globaldsl13}) but we will leave integration of syntax extensions with semantic extensions as future work.
-\section{From Extensible Compilers to Extensible Languages}\label{evolution}
-\begin{figure}[t]
-\small
-$$\begin{array}{rccl}	
-\textbf{programs} & \rho & ::= & \pfam{\familyDf}{\progsort}  \\
- & & \pipe &  \pdef{t}{\kappa}{\tau}{\progsort} \pipe e\\
-\text{primitive ops}		&	\theta	&	::= &	\tops{op}{\kappaidx}{i}{a}{\tau} \pipe 
-												\topp{\theta}{\theta}\\
-\\
-\textbf{external terms} 				&	e	&	::=	&	\evar{x} \pipe 
-														\elam{\evar{x}}{\tau}{e} \pipe 
-														\eop{Fam}{op}{
-															\taui
-														}{
-  												    		\splat{e}{1}{n}
-														} \\
-									& 		&		& 	\\
-
-\hspace{-5pt}\textbf{type-level terms} 	& \tau 	& ::= 	& 	\tvar{t} \pipe 
-														\tlam{t}{\kappa}{\tau} \pipe 
-														\tapp{\tau_1}{\tau_2}\pipe
-														\iintlit \pipe \iop{\tau_1}{\tau_2} \pipe \tstr{str} \\
-									    & &  \pipe & 
-														\tnil{\kappa} \pipe \tcons{\tau_1}{\tau_2} \pipe 
-									                     \tfold{\tau_1}{\tau_2}{h}{t}{r}{\tau_3}
-														\\
-
-	 			& 		& \pipe	& 	 \tunit \pipe 
-														\tpair{\tau_{1}}{\tau_{2}} \pipe 
-														\tfst{\tau} \pipe 
-														\tsnd{\tau} 
-														\\
-\text{equality}  & & \pipe & 					\tifeq{\tau_{1}}{\tau_{2}}{\kappa}{\tau_{3}}{\tau_{4}} 
-														\\
-
-\text{types} 						& 		& \pipe	& 	\ttypestd \\
- & & \pipe &  \tfamcase{\tau}{Tycon}{x}{\tau_1}{\tau_2}\\
-						%				& & \pipe & \tfamcase{\tau}{Fam}{x}{\tau_1}{\tau_2}\\
-
-\text{denotations} 				& 		 & 	\pipe	&	\tden{\tauiterm}{\tautype} \pipe \terr \\
- & & \pipe &  \tdencase{\tau}{x}{t}{\tau_1}{\tau_2}\\
- %& & \pipe & 
-%														\tdencase{\tau}{y}{x}{\tau_1}{\tau_2}
+%\section{From Extensible Compilers to Extensible Languages}\label{evolution}
+%\begin{figure}[t]
+%\small
+%$$\begin{array}{rccl}	
+%\textbf{programs} & \rho & ::= & \pfam{\familyDf}{\progsort}  \\
+% & & \pipe &  \pdef{t}{\kappa}{\tau}{\progsort} \pipe e\\
+%\text{primitive ops}		&	\theta	&	::= &	\tops{op}{\kappaidx}{i}{a}{\tau} \pipe 
+%												\topp{\theta}{\theta}\\
+%\\
+%\textbf{external terms} 				&	e	&	::=	&	\evar{x} \pipe 
+%														\elam{\evar{x}}{\tau}{e} \pipe 
+%														\eop{Fam}{op}{
+%															\taui
+%														}{
+%  												    		\splat{e}{1}{n}
+%														} \\
+%									& 		&		& 	\\
+%							
+%\hspace{-5pt}\textbf{type-level terms} 	& \tau 	& ::= 	& 	\tvar{t} \pipe 
+%														\tlam{t}{\kappa}{\tau} \pipe 
+%														\tapp{\tau_1}{\tau_2}\pipe
+%														\iintlit \pipe \iop{\tau_1}{\tau_2} \pipe \tstr{str} \\
+%									    & &  \pipe & 
+%														\tnil{\kappa} \pipe \tcons{\tau_1}{\tau_2} \pipe 
+%									                     \tfold{\tau_1}{\tau_2}{h}{t}{r}{\tau_3}
 %														\\
-
-\text{reified IL}		&		&	\pipe	&	\titerm{\gamma} \pipe \titype{\sigma} \\
-
-												\\
-\textbf{kinds} 					& \kappa	&	::=	&	\karrow{\kappa_1}{\kappa_2} \pipe \klist{\kappa} \pipe \dint \pipe
-											    \kstr \pipe
-												\kunit \pipe 
-												\kpair{\kappa_{1}}{\kappa_{2}} \\
-	& & \pipe &  
-												\kTypeBlur \pipe \kDen \pipe 
-												\kIType \pipe \kITerm
-												\\
-\\												
-%\textbf{ops signature}			& \Theta	&	::=	&	\kOpEmpty \pipe \kOp{\Theta}{op}{\kappai}\\
-%											 							&		&		&	\\
-\textbf{internal terms} 				& 	\gamma	&	::=	&	\evar{x} \pipe 
-												\ilam{\evar{x}}{\sigma}{\gamma} \pipe 
-												\iapp{\gamma_{1}}{\gamma_{2}} \pipe
-												\ifix{\evar{f}}{\sigma}{\gamma} \\
-									& & \pipe & 
-												\ipair{\gamma_{1}}{\gamma_{2}} \pipe 
-												\ifst{\gamma} \pipe
-												\isnd{\gamma}  
-												\\
-							& 		& 	\pipe	& 
-												\iintlit \pipe \iop{\gamma_{1}}{\gamma_{2}} \pipe \iIfEq{\gamma_{1}}{\gamma_{2}}{\dint}{\gamma_{3}}{\gamma_{4}}  
-												\\
- &  & \pipe & \tvalof{\tau_1}{\tau_2} \pipe \iup{\tau} \\
-%\text{deabstracted}& \gamma & ::= & \mathcal{G}[\gamma, \sigma]\\
-\textbf{internal types}			&	\sigma	&	::=	&    \darrow{\sigma_1}{\sigma_2} \pipe
-												\dint \pipe
-												\dpair{\sigma_1}{\sigma_2} \pipe
-												\trepof{\tau} \pipe \dup{\tau}\\
-\end{array}$$
-%\vspace{-10pt}
-\caption{\small Syntax of \atlam.
+%												
+%	 			& 		& \pipe	& 	 \tunit \pipe 
+%														\tpair{\tau_{1}}{\tau_{2}} \pipe 
+%														\tfst{\tau} \pipe 
+%														\tsnd{\tau} 
+%														\\
+%\text{equality}  & & \pipe & 					\tifeq{\tau_{1}}{\tau_{2}}{\kappa}{\tau_{3}}{\tau_{4}} 
+%														\\														
+%
+%\text{types} 						& 		& \pipe	& 	\ttypestd \\
+% & & \pipe &  \tfamcase{\tau}{Tycon}{x}{\tau_1}{\tau_2}\\
+%						%				& & \pipe & \tfamcase{\tau}{Fam}{x}{\tau_1}{\tau_2}\\
+%																								
+%\text{denotations} 				& 		 & 	\pipe	&	\tden{\tauiterm}{\tautype} \pipe \terr \\
+% & & \pipe &  \tdencase{\tau}{x}{t}{\tau_1}{\tau_2}\\
+% %& & \pipe & 
+%%														\tdencase{\tau}{y}{x}{\tau_1}{\tau_2}
+%%														 \\
+%
+%\text{reified IL}		&		&	\pipe	&	\titerm{\gamma} \pipe \titype{\sigma} \\
+%
+%												\\
+%\textbf{kinds} 					& \kappa	&	::=	&	\karrow{\kappa_1}{\kappa_2} \pipe \klist{\kappa} \pipe \dint \pipe
+%											    \kstr \pipe
+%												\kunit \pipe 
+%												\kpair{\kappa_{1}}{\kappa_{2}} \\
+%	& & \pipe &  
+%												\kTypeBlur \pipe \kDen \pipe 
+%												\kIType \pipe \kITerm
+%												\\
+%\\												
+%%\textbf{ops signature}			& \Theta	&	::=	&	\kOpEmpty \pipe \kOp{\Theta}{op}{\kappai}\\
+%%											 							&		&		&	\\
+%\textbf{internal terms} 				& 	\gamma	&	::=	&	\evar{x} \pipe 
+%												\ilam{\evar{x}}{\sigma}{\gamma} \pipe 
+%												\iapp{\gamma_{1}}{\gamma_{2}} \pipe
+%												\ifix{\evar{f}}{\sigma}{\gamma} \\
+%									& & \pipe & 
+%												\ipair{\gamma_{1}}{\gamma_{2}} \pipe 
+%												\ifst{\gamma} \pipe
+%												\isnd{\gamma}  
+%												\\
+%							& 		& 	\pipe	& 
+%												\iintlit \pipe \iop{\gamma_{1}}{\gamma_{2}} \pipe \iIfEq{\gamma_{1}}{\gamma_{2}}{\dint}{\gamma_{3}}{\gamma_{4}}  
+%												\\
+% &  & \pipe & \tvalof{\tau_1}{\tau_2} \pipe \iup{\tau} \\
+%%\text{deabstracted}& \gamma & ::= & \mathcal{G}[\gamma, \sigma]\\
+%\textbf{internal types}			&	\sigma	&	::=	&    \darrow{\sigma_1}{\sigma_2} \pipe
+%												\dint \pipe
+%												\dpair{\sigma_1}{\sigma_2} \pipe
+%												\trepof{\tau} \pipe \dup{\tau}\\
+%\end{array}$$
+%%\vspace{-10pt}
+%\caption{\small Syntax of \atlam.
 Variables $x$ are used in expressions and internal terms and are distinct from type-level variables, $\tvar{t}$.
@@
 %The productions related to the internal language are written using generators $\mathcal{G}$ and $\mathcal{S}$ to avoid duplicating the syntax of common terms.
-\label{grammar}}
-\end{figure}
-\begin{figure}[t]
-\small
-\begin{flalign}
-\label{natfam}&\family{Nat}{\kunit}{\\
-\label{z}&\quad\tops{z}{\kunit}{i}{a}{
-	\tapp{\tapp{\tvar{empty}}{\tvar{a}}}{
-		\tden{\titerm{0}}{\ttype{Nat}{\tunit}}
-	}};\\
-\label{s}&\quad\tops{s}{\kunit}{i}{a}{
-	\tapp{
-		\tapp{\tvar{pop\_final}}{\tvar{a}}}{\tlam{x}{\kITerm}{\tlam{t}{\kTypeBlur}{
-			\\
-			&\quad\quad\tapp{\tapp{\tapp{\tvar{check\_type}}{\tvar{t}}}{\ttype{Nat}{\tunit}}}{
-				\tden{\titerm{\iup{\tvar{x}}+1}}{\ttype{Nat}{\tunit}}
-			}
-		}}
-		}
-	};\\
-\label{rec}&\quad\tops{rec}{\kunit}{i}{a}{
-	\tapp{\tapp{\tvar{pop}}{\tvar{a}}}{\tlam{x1}{\kITerm}{\tlam{t1}{\kTypeBlur}{\tlam{a}{\klist{\kDen}}{\\
-	&\quad\quad\tapp{\tapp{\tvar{pop}}{\tvar{a}}}{\tlam{x2}{\kITerm}{\tlam{t2}{\kTypeBlur}{\tlam{a}{\klist{\kDen}}{\\
-	&\quad\quad \tapp{\tapp{\tvar{pop\_final}}{\tvar{a}}}{\tlam{x3}{\kITerm}{\tlam{t3}{\kTypeBlur}{\\
-	&\quad\quad \tapp{\tapp{\tapp{\tvar{check\_type}}{\tvar{t1}}}{\ttype{Nat}{\tunit}}}{(\\
-	&\quad\quad \tapp{\tapp{\tapp{\tvar{check\_type}}{\tvar{t3}}}{\ttype{Arrow}{(\ttype{Nat}{\tunit},\ttype{Arrow}{(\tvar{t2},\tvar{t2})})}}}{\\
-		\label{fix}&\quad\quad \tden{\titerm{\iapp{(\ifix{f}{\darrow{\dint}{\trepof{\tvar{t2}}}}{\ilam{x}{\dint}{\\
-		\label{lastop}&\quad\quad\quad \iIfEq{x}{0}{\dint}{\iup{\tvar{x2}}}{\iapp{\iapp{\iup{\tvar{x3}}}{(x-1)}}{(\iapp{f}{(x-1)})}}}})}{\iup{\tvar{x1}}}}}{\tvar{t2}}
-	)}}
-	}}}}}}}}}}}}
-\\
-&}{XXX}{i}{\titype{\dint}};\\
-\label{nattype}&\pdef{nat}{\kTypeBlur}{\ttype{Nat}{\tunit}}{\\
-\label{plus}&(\elam{plus}{\ttype{Arrow}{(\tvar{nat}, \ttype{Arrow}{(\tvar{nat}, \tvar{nat}))}}}{
-	\elam{two}{\tvar{nat}}{\\
-\label{ap}	&\quad\quad \eopapp{\eopapp{plus}{two}}{two}}})
-}\\
-\label{add}&\quad (\elam{x}{\tvar{nat}}{\elam{y}{\tvar{nat}}{
-	\eop{Nat}{rec}{\tunit}{x; y; \elam{p}{\tvar{nat}}{\elam{r}{\tvar{nat}}{
-	\eop{Nat}{s}{\tunit}{r}
-	}}}}})\\
-\label{two}&\quad  \eop{Nat}{s}{\tunit}{\eop{Nat}{s}{\tunit}{\eop{Nat}{z}{\tunit}{ }}}
-%{\\&\elet{two}{\tvar{nat}}{\eop{Nat}{s}{\tunit}{\eop{Nat}{s}{\tunit}{\eop{Nat}{z}{\tunit}{ }}}}{\eapp{\eapp{plus}{two}}{two}}}
-%}
-%\elam{x}{\tvar{nat}}{\elam{y}{\tvar{nat}}{\\
-	%&\quad \eop{Nat}{rec}{\tunit}{x; y; \elam{p}{\tvar{nat}}{\elam{r}{\tvar{nat}}{
-	%\eop{Nat}{s}{\tunit}{r}
-	%}}}}
-\end{flalign}
-\caption{G\"odel's T in \atlam, used to calculate 2+2.
+%\label{grammar}}
+%\end{figure}
+%\begin{figure}[t]
+%\small
+%\begin{flalign}
+%\label{natfam}&\family{Nat}{\kunit}{\\
+%\label{z}&\quad\tops{z}{\kunit}{i}{a}{
+%	\tapp{\tapp{\tvar{empty}}{\tvar{a}}}{
+%		\tden{\titerm{0}}{\ttype{Nat}{\tunit}}
+%	}};\\
+%\label{s}&\quad\tops{s}{\kunit}{i}{a}{
+%	\tapp{
+%		\tapp{\tvar{pop\_final}}{\tvar{a}}}{\tlam{x}{\kITerm}{\tlam{t}{\kTypeBlur}{
+%			\\
+%			&\quad\quad\tapp{\tapp{\tapp{\tvar{check\_type}}{\tvar{t}}}{\ttype{Nat}{\tunit}}}{
+%				\tden{\titerm{\iup{\tvar{x}}+1}}{\ttype{Nat}{\tunit}}
+%			}
+%		}}
+%		}
+%	};\\
+%\label{rec}&\quad\tops{rec}{\kunit}{i}{a}{
+%	\tapp{\tapp{\tvar{pop}}{\tvar{a}}}{\tlam{x1}{\kITerm}{\tlam{t1}{\kTypeBlur}{\tlam{a}{\klist{\kDen}}{\\
+%	&\quad\quad\tapp{\tapp{\tvar{pop}}{\tvar{a}}}{\tlam{x2}{\kITerm}{\tlam{t2}{\kTypeBlur}{\tlam{a}{\klist{\kDen}}{\\
+%	&\quad\quad \tapp{\tapp{\tvar{pop\_final}}{\tvar{a}}}{\tlam{x3}{\kITerm}{\tlam{t3}{\kTypeBlur}{\\
+%	&\quad\quad \tapp{\tapp{\tapp{\tvar{check\_type}}{\tvar{t1}}}{\ttype{Nat}{\tunit}}}{(\\
+%	&\quad\quad \tapp{\tapp{\tapp{\tvar{check\_type}}{\tvar{t3}}}{\ttype{Arrow}{(\ttype{Nat}{\tunit},\ttype{Arrow}{(\tvar{t2},\tvar{t2})})}}}{\\
+%		\label{fix}&\quad\quad \tden{\titerm{\iapp{(\ifix{f}{\darrow{\dint}{\trepof{\tvar{t2}}}}{\ilam{x}{\dint}{\\
+%		\label{lastop}&\quad\quad\quad \iIfEq{x}{0}{\dint}{\iup{\tvar{x2}}}{\iapp{\iapp{\iup{\tvar{x3}}}{(x-1)}}{(\iapp{f}{(x-1)})}}}})}{\iup{\tvar{x1}}}}}{\tvar{t2}}
+%	)}}
+%	}}}}}}}}}}}}
+%\\
+%&}{XXX}{i}{\titype{\dint}};\\
+%\label{nattype}&\pdef{nat}{\kTypeBlur}{\ttype{Nat}{\tunit}}{\\
+%\label{plus}&(\elam{plus}{\ttype{Arrow}{(\tvar{nat}, \ttype{Arrow}{(\tvar{nat}, \tvar{nat}))}}}{
+%	\elam{two}{\tvar{nat}}{\\
+%\label{ap}	&\quad\quad \eopapp{\eopapp{plus}{two}}{two}}})
+%}\\
+%\label{add}&\quad (\elam{x}{\tvar{nat}}{\elam{y}{\tvar{nat}}{
+%	\eop{Nat}{rec}{\tunit}{x; y; \elam{p}{\tvar{nat}}{\elam{r}{\tvar{nat}}{
+%	\eop{Nat}{s}{\tunit}{r}
+%	}}}}})\\
+%\label{two}&\quad  \eop{Nat}{s}{\tunit}{\eop{Nat}{s}{\tunit}{\eop{Nat}{z}{\tunit}{ }}}
+%%{\\&\elet{two}{\tvar{nat}}{\eop{Nat}{s}{\tunit}{\eop{Nat}{s}{\tunit}{\eop{Nat}{z}{\tunit}{ }}}}{\eapp{\eapp{plus}{two}}{two}}}
+%%}
+%%\elam{x}{\tvar{nat}}{\elam{y}{\tvar{nat}}{\\
+%	%&\quad \eop{Nat}{rec}{\tunit}{x; y; \elam{p}{\tvar{nat}}{\elam{r}{\tvar{nat}}{
+%	%\eop{Nat}{s}{\tunit}{r}
+%	%}}}}
+%\end{flalign}
+%\caption{G\"odel's T in \atlam, used to calculate 2+2.
 Helper functions for working with lists ($\tvar{empty}$, $\tvar{pop}$, $\tvar{pop\_final}$) and types ($\tvar{check\_type}$),  described below, are given in the appendix.
 We will refer to this program as $\rho_{\text{nat}}$ in the text.}
-\label{example}
-\end{figure}
-
-To understand the genesis of our internal extension mechanism, it is helpful to begin by considering why most implementations of programming languages cannot even be  externally extended. 
-Let us consider, as a simple example, an implementation of G\"odel's T, a typed lambda calculus with recursion on primitive natural numbers (see Appendix). 
-A compiler for this language written using a functional language will invariably represent the primitive type families and operators using {closed} inductive datatypes. 
-For example, a simple implementation in Standard ML may be based around these datatypes:
-\begin{lstlisting}
-  datatype Type = Nat | Arrow of Type * Type
-  datatype Exp = Var of var 
-               | Lam of var * Type * Exp | Ap of Exp * Exp 
-               | Z | S of Exp | Natrec of Exp * Exp * Exp
-\end{lstlisting}
-
-The logic governing typechecking and translation to a suitable intermediate language (for subsequent optimization and compilation by some back-end) will proceed by exhaustive case analysis over the constructors of \lstinline{Exp}.
-
-In an object-oriented implementation of Godel's T, we might instead encode types and operators as subclasses of abstract classes \lstinline{Type} and \lstinline{Exp}.
+%\label{example}
+%\end{figure}
+%
+%To understand the genesis of our internal extension mechanism, it is helpful to begin by considering why most implementations of programming languages cannot even be  externally extended. 
+%Let us consider, as a simple example, an implementation of G\"odel's T, a typed lambda calculus with recursion on primitive natural numbers (see Appendix). 
+%A compiler for this language written using a functional language will invariably represent the primitive type families and operators using {closed} inductive datatypes. 
+%For example, a simple implementation in Standard ML may be based around these datatypes:
+%\begin{lstlisting}
+%  datatype Type = Nat | Arrow of Type * Type
+%  datatype Exp = Var of var 
+%               | Lam of var * Type * Exp | Ap of Exp * Exp 
+%               | Z | S of Exp | Natrec of Exp * Exp * Exp
+%\end{lstlisting}
+%
+%The logic governing typechecking and translation to a suitable intermediate language (for subsequent optimization and compilation by some back-end) will proceed by exhaustive case analysis over the constructors of \lstinline{Exp}.
+%
+%In an object-oriented implementation of Godel's T, we might instead encode types and operators as subclasses of abstract classes \lstinline{Type} and \lstinline{Exp}.
 Typechecking and translation will proceed by the ubiquitous \emph{visitor pattern}  by dispatching against a fixed collection of {known} subclasses of \lstinline{Exp}.
-
-In either case, we encounter the same basic issue: there is no way to modularly add new primitive type families and operators and implement their associated typechecking and translation logic. 
-%This issue is related to the widely-discussed \emph{expression problem} (in a restricted sense -- we do not consider adding new functions beyond typechecking and translation here, only adding logic to these) \cite{wadler-expression}.
-
-A number of language mechanisms have been proposed that allow new cases to be added to datatypes and the functions that operate over them in a modular manner. 
-In functional languages, we might use \emph{open datatypes}.
+%
+%In either case, we encounter the same basic issue: there is no way to modularly add new primitive type families and operators and implement their associated typechecking and translation logic. 
+%%This issue is related to the widely-discussed \emph{expression problem} (in a restricted sense -- we do not consider adding new functions beyond typechecking and translation here, only adding logic to these) \cite{wadler-expression}.
+%
+%A number of language mechanisms have been proposed that allow new cases to be added to datatypes and the functions that operate over them in a modular manner. 
+%In functional languages, we might use \emph{open datatypes}.
 For example, if we wish to extend G\"odel's T with product types and we have written our compiler in a language supporting open inductive datatypes, it might be possible to add new cases like this:
-\begin{lstlisting}
-  newcase Prod of Type * Type extends Type
-  newcase Pair of Exp * Exp extends Exp    (* Intro *)
-  newcase PrL of Exp extends Exp           (* Elim Left *)
-  newcase PrR of Exp extends Exp           (* Elim Right *)
-\end{lstlisting}
-
-The logic for functionality like typechecking and translation could then be implemented for only these new cases.
+%\begin{lstlisting}
+%  newcase Prod of Type * Type extends Type
+%  newcase Pair of Exp * Exp extends Exp    (* Intro *)
+%  newcase PrL of Exp extends Exp           (* Elim Left *)
+%  newcase PrR of Exp extends Exp           (* Elim Right *)
+%\end{lstlisting}
+%
+%The logic for functionality like typechecking and translation could then be implemented for only these new cases.
 For example, the \lstinline{typeof} function that assigns a type to an expression could be extended like so:
-\begin{lstlisting}
-  typeof PrL(e) = case typeof e of 
-      Prod(t1, _) => t1 
-    | _ => raise TypeError("<appropriate error message>")
-\end{lstlisting}
-
-If we allowed users to define new modules containing definitions like these and link them into our compiler, we will have succeeded in creating an externally-extensible compiler, albeit one where safety is not guaranteed (we will return to this point shortly).
+%\begin{lstlisting}
+%  typeof PrL(e) = case typeof e of 
+%      Prod(t1, _) => t1 
+%    | _ => raise TypeError("<appropriate error message>")
+%\end{lstlisting}
+%
+%If we allowed users to define new modules containing definitions like these and link them into our compiler, we will have succeeded in creating an externally-extensible compiler, albeit one where safety is not guaranteed (we will return to this point shortly).
 We have not, however, created an extensible programming language, for two reasons.
@@
 Second, other compilers for the same language will not necessarily support the same extensions.
-If our newly-introduced constructs are exposed at a library's  interface boundary, clients using different compilers face the same problems with interoperability that those using different languages face.
+%If our newly-introduced constructs are exposed at a library's  interface boundary, clients using different compilers face the same problems with interoperability that those using different languages face.
 That is, {extending a language by extending a single compiler for it is morally equivalent to creating a new language}.
 Several prominent language ecosystems today are in a state where a prominent compiler has introduced or enabled the introduction of extensions that many libraries have come to rely on, including the Glasgow Haskell Compiler, SML/NJ and the GNU compilers for C and C++.
-
-A more appropriate and useful place for extensions like this is directly within libraries, alongside abstractions that can be adequately implemented in terms of existing primitive abstractions.
+%
+%A more appropriate and useful place for extensions like this is directly within libraries, alongside abstractions that can be adequately implemented in terms of existing primitive abstractions.
 To enable this, the language must allow for the introduction new primitive type families, like \lstinline{Prod}, operators, like \lstinline{Pair}, \lstinline{PrL} and \lstinline{PrR}, and associated typechecking and translation logic.
@@
 Because this mechanism is {language-internal}, all compilers must support it to satisfy the language specification.
-
-Statically-typed languages typically make a distinction between \emph{expressions}, which describe run-time computations, and type-level constructs like types, type aliases and datatype declarations.
+%
+%Statically-typed languages typically make a distinction between \emph{expressions}, which describe run-time computations, and type-level constructs like types, type aliases and datatype declarations.
 The design described above suggests we may now need to add another layer to our language, an {extension language}, where extensions can be declared and implemented.
@@
 This is precisely the structure that a distinct extension layer would have, and so it is quite natural to unify the two, as we will show in this work.
-
-\section{\atlam}\label{atlam}
-In this section, we will develop a core calculus, called @$\lambda$ for the ``actively-typed lambda calculus'', by way of a semantics and a simple example, and discuss how it addresses the safety concerns that arise.
-\subsection{Overview}
-The grammar of \atlam~is shown in Figure \ref{grammar}. 
-A program, $\rho$, consists of a series of declarations followed by an expression.
+%
+%\section{\atlam}\label{atlam}
+%In this section, we will develop a core calculus, called @$\lambda$ for the ``actively-typed lambda calculus'', by way of a semantics and a simple example, and discuss how it addresses the safety concerns that arise.
+%\subsection{Overview}
+%The grammar of \atlam~is shown in Figure \ref{grammar}. 
+%A program, $\rho$, consists of a series of declarations followed by an expression.
 Declarations can be either bindings of type-level terms to type-level variables using \textsf{def} or a primitive type family declared using \textsf{family}, which can contain implementations of one or more operator families, $\theta$.
 Expressions, $e$, can be either variables, lambdas, or applications of operators.
-%, and are ultimately given meaning by translation to a typed internal language, with terms $\gamma$ and types $\sigma$. 
-%This language has been chosen, for simplicity, to be a variant of Plotkin's PCF with primitive integers and products, but in practice would include other constructs consistent with its role as a high-level intermediate language.
-
-The language is structured as a simply-typed lambda calculus with simply-kinded type-level computation. Kinds, $\kappa$, classify type-level terms, $\tau$. 
-Types are type-level values of kind $\star$ (following System $F_{\omega}$) and classify expressions.
+%%, and are ultimately given meaning by translation to a typed internal language, with terms $\gamma$ and types $\sigma$. 
+%%This language has been chosen, for simplicity, to be a variant of Plotkin's PCF with primitive integers and products, but in practice would include other constructs consistent with its role as a high-level intermediate language.
+%
+%The language is structured as a simply-typed lambda calculus with simply-kinded type-level computation. Kinds, $\kappa$, classify type-level terms, $\tau$. 
+%Types are type-level values of kind $\star$ (following System $F_{\omega}$) and classify expressions.
 The type-level language also includes other kinds of terms: type-level functions, lists (required by our mechanism), integers, strings and products for the sake of our examples (see Sec.
 \ref{safety} for a discussion on other acceptable kinds of type-level data) and constructs for developing extensions -- denotations and reified internal terms and types -- which we will discuss in the sections below.
-
-All expressions are given meaning by translation to a typed internal language.
+%
+%All expressions are given meaning by translation to a typed internal language.
 This language has been chosen, for simplicity, to be a variant of Plotkin's PCF with primitive integers and products, but in practice would include other constructs consistent with its role as a high-level intermediate language.
 The grammar of internal terms, $\gamma$, and internal types, $\sigma$, also includes special forms containing type-level terms; these are used for developing extensions and during compilation and will be erased before compilation ends.
-\subsection{Example: G\"odel's T as an Active Type Family}
-To make our explanation of each of the constructs in the calculus concrete, we will work through an example showing how to introduce primitive natural numbers with bounded recursion in the style of G\"odel's T \cite{pfpl}.
+%\subsection{Example: G\"odel's T as an Active Type Family}
+%To make our explanation of each of the constructs in the calculus concrete, we will work through an example showing how to introduce primitive natural numbers with bounded recursion in the style of G\"odel's T \cite{pfpl}.
 These will be implemented internally as integers (that is, internal terms of internal type $\dint$).
@@
 We will refer back to this figure as we describe each construct below.
-
-\subsection{Indexed Type Families and Types}\label{families}
-The syntactic form $\familyDf$ declares a new primitive type family named $\fvar{Fam}$ indexed by type-level values of kind $\kappaidx$ with representation schema $\tvar{i}.\taurep$ and operators $\theta$.
+%
+%\subsection{Indexed Type Families and Types}\label{families}
+%The syntactic form $\familyDf$ declares a new primitive type family named $\fvar{Fam}$ indexed by type-level values of kind $\kappaidx$ with representation schema $\tvar{i}.\taurep$ and operators $\theta$.
 The purpose of the representation schema and of associating of operators directly with types will be explained below.
-
-Declaring a type family in this way is a language-internal analog to adding a new constructor to the compiler-internal datatype \lstinline{Type}, as suggested in Sec. \ref{evolution}. 
-The index represents the data associated with this constructor.
+%
+%Declaring a type family in this way is a language-internal analog to adding a new constructor to the compiler-internal datatype \lstinline{Type}, as suggested in Sec. \ref{evolution}. 
+%The index represents the data associated with this constructor.
 A type (that is, a type-level term of kind $\kTypeBlur$) is constructed by naming a family in scope and providing a type-level term of the appropriate kind as an index.
 A base type like $\tvar{nat}$ can be thought of as being the only type in the family $\fvar{Nat}$ trivially indexed by the unit value, of kind $\kunit$, while families like $\fvar{Ntuple}$ might be indexed by a list of types, having kind $\klist{\kTypeBlur}$.
- For example, $\ttype{Ntuple}{\tcons{\tvar{nat}}{\tcons{\tvar{nat}}{\tnil{\kTypeBlur}}}}$ might be the type of a pair of natural numbers.
+% For example, $\ttype{Ntuple}{\tcons{\tvar{nat}}{\tcons{\tvar{nat}}{\tnil{\kTypeBlur}}}}$ might be the type of a pair of natural numbers.
 Given a type, its family can be \textsf{case} analyzed to extract the value of its index.
@@
 The main  consequence of this restriction is that indices cannot contain type-level functions.
-
-\subsection{Representations and Representation Schemas}
-As we will discuss further below, it is important that all expressions classified by a type compile to consistently-typed internal terms.
+%
+%\subsection{Representations and Representation Schemas}
+%As we will discuss further below, it is important that all expressions classified by a type compile to consistently-typed internal terms.
 For this reason, we require that every type have a single internal type associated with it, called its \emph{representation}.
@@
 Internal types, $\sigma$, are reified as type-level terms of kind $\kIType$ using the introductory form $\titype{\sigma}$.
-
-\subsection{Indexed Operator Families and Denotations}\label{operators}
-Type families are also equiped with a collection of primitive operator families.
+%
+%\subsection{Indexed Operator Families and Denotations}\label{operators}
+%Type families are also equiped with a collection of primitive operator families.
 An operator family named $\opvar{op}$ is declared using the form $\tops{op}{\kappaidx}{i}{a}{\tauop}$.
@@
 \ref{examples}).
-
-To apply an operator, the grammar provides a uniform form of expression: $\eop{Fam}{op}{\tauidx}{e_1; \ldots; e_n}$, where $n \geq 0$.
+%
+%To apply an operator, the grammar provides a uniform form of expression: $\eop{Fam}{op}{\tauidx}{e_1; \ldots; e_n}$, where $n \geq 0$.
 The typechecking and translation of an expression of this form is controlled by the term $\tauop$ in the operator's declaration.
 This term must evaluate to a \emph{denotation}, which is a type-level value of kind $\kDen$, when given the operator index, $\tauidx$, and a list constructed from the denotations recursively assigned to each argument, $e_1$ through $e_n$.
-
-There are two forms of denotations that an expression can be assigned.
+%
+%There are two forms of denotations that an expression can be assigned.
 A \emph{valid denotation} has the form $\tden{\tauiterm}{\tautype}$, where $\tauiterm$ is the \emph{translation} of the expression to an internal term and $\tautype$ is the type it has been assigned.
@@
 %It is important to note that there are no elimination form for reified terms or types.
-%(indeed, in \atlam, terms are never  syntactically analyzed directly by extensions, unlike macro systems and other forms of term rewriting; see Sec. \ref{related-work})
-If a type error is detected by an operator, the \emph{error denotation}, $\terr$, is returned instead of a valid denotation.
+%%(indeed, in \atlam, terms are never  syntactically analyzed directly by extensions, unlike macro systems and other forms of term rewriting; see Sec. \ref{related-work})
+%If a type error is detected by an operator, the \emph{error denotation}, $\terr$, is returned instead of a valid denotation.
 In a practical implementation, a specialized error message and other diagnostic information would be provided when returning $\terr$, but we omit such details for simplicity.
 Terms of kind $\kDen$ can be \textsf{case} analyzed to determine if they are valid or errors, and if valid, to extract the translation and type.
-
-In the example in Fig.
+%
+%In the example in Fig.
 \ref{example}, the operator $\opvar{z}$ checks that no arguments were passed in using the simple helper function $\tvar{empty}$.
@@
 \ref{repcon}.
-
-% The operator index, $\tauidx$, and .
-
-\subsection{Functions, Variables and Arrow Types}
-
-The recursor operator, $\opvar{rec}$, proceeds similarly, extracting the translations and types of each of its three arguments.
+%
+%% The operator index, $\tauidx$, and . 
+%
+%\subsection{Functions, Variables and Arrow Types}
+%
+%The recursor operator, $\opvar{rec}$, proceeds similarly, extracting the translations and types of each of its three arguments.
 Of note, however, is how it handles the third argument, which binds two variables (the predecessor and the result of recursing on it).
@@
 The details are straightforward and given in the appendix.
-
-\subsection{Compilation}
-\begin{figure}[t]
-\small
-\begin{mathpar}
-\inferrule{
-	\overbrace{\progOK{\emptyctx}{\fvarCtx_0}{\rho}}^{\text{\normalsize Kind Checking}}\\
-	\overbrace{\pcompiles{\fvalCtx_0}{\rho}{\gamma}}^{\text{\normalsize Active Typechecking and Translation}}
-}{
-	\ptcc{\rho}{\gamma}
-}
-\end{mathpar}
-\caption{\small Central Compilation Judgement of \atlam.}
-\label{ccj}
-\end{figure}
-The \emph{central compilation judgement}, shown in Figure \ref{ccj}, captures the two phases of compilation: kind checking and active typechecking and translation. 
-The first phase ensures that all type-level terms in the program are well-kinded and that all expressions and internal terms are closed.
+%
+%\subsection{Compilation}
+%\begin{figure}[t]
+%\small
+%\begin{mathpar}
+%\inferrule{
+%	\overbrace{\progOK{\emptyctx}{\fvarCtx_0}{\rho}}^{\text{\normalsize Kind Checking}}\\
+%	\overbrace{\pcompiles{\fvalCtx_0}{\rho}{\gamma}}^{\text{\normalsize Active Typechecking and Translation}}
+%}{
+%	\ptcc{\rho}{\gamma}
+%}
+%\end{mathpar}
+%\caption{\small Central Compilation Judgement of \atlam.}
+%\label{ccj}
+%\end{figure}
+%The \emph{central compilation judgement}, shown in Figure \ref{ccj}, captures the two phases of compilation: kind checking and active typechecking and translation. 
+%The first phase ensures that all type-level terms in the program are well-kinded and that all expressions and internal terms are closed.
 The kinding rules for programs are given in Figure \ref{kindprog}, and they rely on the kinding rules for type-level terms given in Figure \ref{tlkind}.
@@
 \ref{example} is well-kinded.
-% As we will see, the kind checking phase ensures that many kinds of errors are ruled out and that this process will not ``get stuck'' (see Sec.
+%% As we will see, the kind checking phase ensures that many kinds of errors are ruled out and that this process will not ``get stuck'' (see Sec.
 \ref{safety}).
@@
 \ref{tleval}.
-\begin{figure}[t]
-\small
-$\fbox{\inferrule{}{\progOKX{\progsort}}}$
-~~~$\tvarCtx ::= \emptyctx \pipe \tvarCtxX{t}{\kappa}$~~~
-$\fCtx ::= \Sigma_0	 \pipe \fvarCtxX$
-\begin{mathpar}
-\inferrule[family-kinding]{
-	\fvar{Fam} \notin \text{dom}(\fvarCtx)\\
-	\kEq{\kappaidx}\\
-	\opType{\tvarCtx}{\fvarCtxX}{\theta}{\Theta}\\\\
-	\tKind{\tvarCtx, \tvar{i}:{\kappaidx}}{\fvarCtxX}{\tau}{\kIType}\\
-	\progOK{\tvarCtx}{\fvarCtxX}{\rho}
-}{
-	\progOKX{\pfam{\familyDf}{\rho}}
-}
-
-\inferrule[def-kinding]{
-	\tKindX{\tau}{\kappa}\\
-	\progOK{\tvarCtxX{t}{\kappa}}{\fvarCtx}{\rho}
-}{
-	\progOKX{\pdef{t}{\kappa}{\tau}{\rho}}
-}
-
-\inferrule[exp-kinding]{
-	\exprOK{\tvarCtx}{\emptyctx}{\fvarCtx}{e}
-}{
-	\progOKX{e}
-}
-\end{mathpar}
-$\fbox{$\tKindX{\theta}{\Theta}$}$
-~~~$\Theta ::= \kOpS{op}{\kappaidx} \pipe \Theta, \Theta$~~~
-\begin{mathpar}
-\inferrule[op-kinding]{
-	\tKind{\tvarCtx, \tOfKind{\tvar{i}}{\kappai}, \tOfKind{\tvar{a}}{\klist{\kDen}}}{\fvarCtx}{\tau}{\kDen}
-}{
-	\opType{\tvarCtx}{\fvarCtx}{{\tops{op}{\kappaidx}{i}{a}{\tau}}}{
-	\kOpS{op}{\kappaidx}}
-}
-
-\inferrule[ops-kinding]{
-	\tKindX{\theta_1}{\Theta_1}\\
-	\tKindX{\theta_2}{\Theta_2}\\\\
-	\text{dom}(\theta_1) \cap \text{dom}(\theta_2) = \emptyset
-}{
-	\tKindX{\theta_1; \theta_2}{\Theta_1, \Theta_2}
-}
-\end{mathpar}
-$\fbox{$\exprOKX{e}$}$
-~~~$\itvarCtx ::= \emptyctx \pipe \itvarCtx, \evar{x}$
-\begin{mathpar}
-\inferrule[e-var-kinding]{ }{
-	\exprOK{\tvarCtx}{\itvarCtx, \evar{x}}{\fvarCtx}{\evar{x}}
-}
-
-\inferrule[e-lam-kinding]{
-	\tKindX{\tau}{\kTypeBlur}\\
-	\exprOK{\tvarCtx}{\eivarCtxX{x}}{\fvarCtx}{e}
-}{
-	\exprOKX{\elam{\evar{x}}{\tau}{e}}
-}
-
-\inferrule[e-op-kinding]{
-	\fvarOfType{Fam}{\kappaidx}{\Theta} \in \fvarCtx\\
-	\kOpS{op}{\kappai} \in \Theta\\
-	\tKindX{\taui}{\kappai}\\\\
-	\exprOKX{e_1}\\
-	\cdots\\
-	\exprOKX{e_n}
-}{
-	\exprOKX{\eop{Fam}{op}{\taui}{\splat{e}{1}{n}}}
-}
-\end{mathpar}
-\caption{\small Kinding for programs. Variable contexts $\tvarCtx$ and $\itvarCtx$ obey standard structural properties. Kinding rules for type-level terms are given in Figure \ref{tlkind}.}
-\label{kindprog}
-\vspace{-10pt}
-\end{figure}
-%\subsection{Active Typechecking and Translation}
-\begin{figure}[t]
-\small
-$\fbox{\inferrule{}{\pcompilesX{\rho}}}$
-~~~$\fvalCtx ::= \fvalCtx_0 \pipe \fvalCtxX{\fvalDf}$
-\begin{mathpar}
-\inferrule[att-fam]{
-	\pcompiles{\fvalCtxX{\fvalDf}}{\rho}{\gamma}
-}{
-	\pcompilesX{\pfam{\familyDf}{\rho}}
-}
-
-\inferrule[att-def]{
-	\tEvalX{\tau}{\tau'}\\
-	\pcompilesX{\subst{\tau'}{\tvar{t}}{\rho}}
-}{
-	\pcompilesX{\pdef{t}{\kappa}{\tau}{\rho}}
-}
-
-\inferrule[att-exp]{
-	\ecompiles{\emptyctx}{\fvalCtx}{e}{\tau}{\gabs}\\
-	\eraseX{\gabs}{\gamma}
-}{
-	\pcompilesX{e}
-}
-\end{mathpar}
-$\fbox{\inferrule{}{\ecompilesX{e}{\tau}{\gabs}}}$
-~~~$\etCtx ::= \emptyctx \pipe \etCtxX{x}{\tau}$
-\begin{mathpar}
-\inferrule[att-var]{ }{
-	\ecompiles{\etCtxX{x}{\tau}}{\fvalCtx}{\evar{x}}{\tau}{\evar{x}}
-}
-
-\inferrule[att-lam]{
-	\tEvalX{\tau_1}{\tau_1'}\\
-	%\delfromtau{$\Xi_0$}{\fvalCtx}{\tau_1'}{\sabs}\\\\
-	\ecompiles{\etCtxX{x}{\tau_1'}}{\fvalCtx}{e}{\tau_2}{\gabs}
-}{
-	\ecompiles{\etCtx}{\fvalCtx}{\elam{\evar{x}}{\tau_1}{e}}{\ttype{Arrow}{\tpair{\tau_1'}{\tau_2}}}{\ilam{\evar{x}}{\trepof{\tau_1'}}{\gabs}}
-}
-
-\inferrule[att-op]{
-	\fvalDf \in \fvalCtx\\
-	\tops{op}{\kappaidx}{i}{a}{\tauop} \in \theta\\
-	\tEvalX{\tauidx}{\tauidx'}\\\\
-	\ecompiles{\etCtx}{\fvalCtx}{e_1}{\tau_1}{\gamma_1}~~~~
-	\cdots~~~~
-	\ecompiles{\etCtx}{\fvalCtx}{e_n}{\tau_n}{\gamma_n}\\
-	\text{$\begin{array}{c}
-		\left[\text{$\begin{array}{r}
-			\tauidx'/\tvar{i}\\
-			\tden{\titerm{\gamma_1}}{\tau_1} :: \cdots :: \tden{\titerm{\gamma_n}}{\tau_n} :: \tnil{\kDen}/\tvar{a}
-			\end{array}$}\right]\tauop
-		\Downarrow
-		\tden{\titerm{\gabs}}{\tau}
-		\end{array}$}\\
-	%\delfromtau{Fam}{\fvalCtx}{\tau}{\sabs}\\
-	%
-	\ddbar{\Xi_0,\fvar{Fam}}{\fvalCtx}{\trepof{\tau}}{\sabs}\\
-	\eCtxTogCtx{\Xi_0,\fvar{Fam}}{\fvalCtx}{\etCtx}{\gtCtx}\\
-	\checkRC{\gtCtx}{\Xi_0,\fvar{Fam}}{\fvalCtx}{\gabs}{\sabs}
-}{
-	\ecompiles{\etCtx}{\fvalCtx}{\eop{Fam}{op}{\tauidx}{\splat{e}{1}{n}}}{\tau}{\gabs}
-}
-\end{mathpar}
-\caption{\small Active typechecking and translation of programs. Evaluation semantics for type-level terms are given in Fig. \ref{tleval}.}
-\label{att}
-\vspace{-10pt}
-\end{figure}
-The second phase of compilation involves invoking the  logic implemented by user-defined operator families to typecheck and translate the program.
+%\begin{figure}[t]
+%\small
+%$\fbox{\inferrule{}{\progOKX{\progsort}}}$
+%~~~$\tvarCtx ::= \emptyctx \pipe \tvarCtxX{t}{\kappa}$~~~
+%$\fCtx ::= \Sigma_0	 \pipe \fvarCtxX$
+%\begin{mathpar}
+%\inferrule[family-kinding]{
+%	\fvar{Fam} \notin \text{dom}(\fvarCtx)\\
+%	\kEq{\kappaidx}\\
+%	\opType{\tvarCtx}{\fvarCtxX}{\theta}{\Theta}\\\\
+%	\tKind{\tvarCtx, \tvar{i}:{\kappaidx}}{\fvarCtxX}{\tau}{\kIType}\\
+%	\progOK{\tvarCtx}{\fvarCtxX}{\rho}
+%}{
+%	\progOKX{\pfam{\familyDf}{\rho}}
+%}
+%
+%\inferrule[def-kinding]{
+%	\tKindX{\tau}{\kappa}\\
+%	\progOK{\tvarCtxX{t}{\kappa}}{\fvarCtx}{\rho}
+%}{
+%	\progOKX{\pdef{t}{\kappa}{\tau}{\rho}}
+%}
+%
+%\inferrule[exp-kinding]{
+%	\exprOK{\tvarCtx}{\emptyctx}{\fvarCtx}{e}
+%}{
+%	\progOKX{e}
+%}
+%\end{mathpar}
+%$\fbox{$\tKindX{\theta}{\Theta}$}$
+%~~~$\Theta ::= \kOpS{op}{\kappaidx} \pipe \Theta, \Theta$~~~
+%\begin{mathpar}
+%\inferrule[op-kinding]{
+%	\tKind{\tvarCtx, \tOfKind{\tvar{i}}{\kappai}, \tOfKind{\tvar{a}}{\klist{\kDen}}}{\fvarCtx}{\tau}{\kDen}
+%}{
+%	\opType{\tvarCtx}{\fvarCtx}{{\tops{op}{\kappaidx}{i}{a}{\tau}}}{
+%	\kOpS{op}{\kappaidx}}
+%}
+%
+%\inferrule[ops-kinding]{
+%	\tKindX{\theta_1}{\Theta_1}\\
+%	\tKindX{\theta_2}{\Theta_2}\\\\
+%	\text{dom}(\theta_1) \cap \text{dom}(\theta_2) = \emptyset
+%}{
+%	\tKindX{\theta_1; \theta_2}{\Theta_1, \Theta_2}
+%}
+%\end{mathpar}
+%$\fbox{$\exprOKX{e}$}$
+%~~~$\itvarCtx ::= \emptyctx \pipe \itvarCtx, \evar{x}$
+%\begin{mathpar}
+%\inferrule[e-var-kinding]{ }{
+%	\exprOK{\tvarCtx}{\itvarCtx, \evar{x}}{\fvarCtx}{\evar{x}}
+%}
+%
+%\inferrule[e-lam-kinding]{
+%	\tKindX{\tau}{\kTypeBlur}\\
+%	\exprOK{\tvarCtx}{\eivarCtxX{x}}{\fvarCtx}{e}
+%}{
+%	\exprOKX{\elam{\evar{x}}{\tau}{e}}
+%}
+%
+%\inferrule[e-op-kinding]{
+%	\fvarOfType{Fam}{\kappaidx}{\Theta} \in \fvarCtx\\
+%	\kOpS{op}{\kappai} \in \Theta\\
+%	\tKindX{\taui}{\kappai}\\\\
+%	\exprOKX{e_1}\\
+%	\cdots\\
+%	\exprOKX{e_n}
+%}{
+%	\exprOKX{\eop{Fam}{op}{\taui}{\splat{e}{1}{n}}}
+%}
+%\end{mathpar}
+%\caption{\small Kinding for programs. Variable contexts $\tvarCtx$ and $\itvarCtx$ obey standard structural properties. Kinding rules for type-level terms are given in Figure \ref{tlkind}.}
+%\label{kindprog}
+%\vspace{-10pt}
+%\end{figure}
+%%\subsection{Active Typechecking and Translation}
+%\begin{figure}[t]
+%\small
+%$\fbox{\inferrule{}{\pcompilesX{\rho}}}$
+%~~~$\fvalCtx ::= \fvalCtx_0 \pipe \fvalCtxX{\fvalDf}$
+%\begin{mathpar}
+%\inferrule[att-fam]{
+%	\pcompiles{\fvalCtxX{\fvalDf}}{\rho}{\gamma}
+%}{
+%	\pcompilesX{\pfam{\familyDf}{\rho}}
+%}
+%
+%\inferrule[att-def]{
+%	\tEvalX{\tau}{\tau'}\\
+%	\pcompilesX{\subst{\tau'}{\tvar{t}}{\rho}}
+%}{
+%	\pcompilesX{\pdef{t}{\kappa}{\tau}{\rho}}
+%}
+%
+%\inferrule[att-exp]{
+%	\ecompiles{\emptyctx}{\fvalCtx}{e}{\tau}{\gabs}\\
+%	\eraseX{\gabs}{\gamma}
+%}{
+%	\pcompilesX{e}
+%}
+%\end{mathpar}
+%$\fbox{\inferrule{}{\ecompilesX{e}{\tau}{\gabs}}}$
+%~~~$\etCtx ::= \emptyctx \pipe \etCtxX{x}{\tau}$
+%\begin{mathpar}
+%\inferrule[att-var]{ }{
+%	\ecompiles{\etCtxX{x}{\tau}}{\fvalCtx}{\evar{x}}{\tau}{\evar{x}}
+%}
+%
+%\inferrule[att-lam]{
+%	\tEvalX{\tau_1}{\tau_1'}\\
+%	%\delfromtau{$\Xi_0$}{\fvalCtx}{\tau_1'}{\sabs}\\\\
+%	\ecompiles{\etCtxX{x}{\tau_1'}}{\fvalCtx}{e}{\tau_2}{\gabs}
+%}{
+%	\ecompiles{\etCtx}{\fvalCtx}{\elam{\evar{x}}{\tau_1}{e}}{\ttype{Arrow}{\tpair{\tau_1'}{\tau_2}}}{\ilam{\evar{x}}{\trepof{\tau_1'}}{\gabs}}
+%}
+%
+%\inferrule[att-op]{
+%	\fvalDf \in \fvalCtx\\
+%	\tops{op}{\kappaidx}{i}{a}{\tauop} \in \theta\\
+%	\tEvalX{\tauidx}{\tauidx'}\\\\
+%	\ecompiles{\etCtx}{\fvalCtx}{e_1}{\tau_1}{\gamma_1}~~~~
+%	\cdots~~~~
+%	\ecompiles{\etCtx}{\fvalCtx}{e_n}{\tau_n}{\gamma_n}\\
+%	\text{$\begin{array}{c}
+%		\left[\text{$\begin{array}{r}
+%			\tauidx'/\tvar{i}\\
+%			\tden{\titerm{\gamma_1}}{\tau_1} :: \cdots :: \tden{\titerm{\gamma_n}}{\tau_n} :: \tnil{\kDen}/\tvar{a}
+%			\end{array}$}\right]\tauop
+%		\Downarrow
+%		\tden{\titerm{\gabs}}{\tau}
+%		\end{array}$}\\
+%	%\delfromtau{Fam}{\fvalCtx}{\tau}{\sabs}\\
+%	%
+%	\ddbar{\Xi_0,\fvar{Fam}}{\fvalCtx}{\trepof{\tau}}{\sabs}\\
+%	\eCtxTogCtx{\Xi_0,\fvar{Fam}}{\fvalCtx}{\etCtx}{\gtCtx}\\
+%	\checkRC{\gtCtx}{\Xi_0,\fvar{Fam}}{\fvalCtx}{\gabs}{\sabs}
+%}{
+%	\ecompiles{\etCtx}{\fvalCtx}{\eop{Fam}{op}{\tauidx}{\splat{e}{1}{n}}}{\tau}{\gabs}
+%}
+%\end{mathpar}
+%\caption{\small Active typechecking and translation of programs. Evaluation semantics for type-level terms are given in Fig. \ref{tleval}.}
+%\label{att}
+%\vspace{-10pt}
+%\end{figure}
+%The second phase of compilation involves invoking the  logic implemented by user-defined operator families to typecheck and translate the program.
 The rules for this phase are given in Fig.
@@
 This is again a largely unsurprising collection of rules with the exception of \textsc{dencase-eval-valid}, which we will explain below.
-
-\subsection{Abstract Representations and Abstract Internal Types}\label{repcon}
-To typecheck and translate an expression, $e$, the rule \textsc{att-exp} first assigns a type, $\tau$, and an \emph{abstract translation}, $\gabs$, to it, as determined by the judgement $\ecompilesX{e}{\tau}{\gabs}$.
+%
+%\subsection{Abstract Representations and Abstract Internal Types}\label{repcon}
+%To typecheck and translate an expression, $e$, the rule \textsc{att-exp} first assigns a type, $\tau$, and an \emph{abstract translation}, $\gabs$, to it, as determined by the judgement $\ecompilesX{e}{\tau}{\gabs}$.
 It then \emph{deabstracts} this abstract translation to complete the compilation process, as determined by the judgement $\eraseX{\gabs}{\gamma}$.
@@
 If the knowledge that natural numbers are implemented as integers was externally visible, a different extension could introduce an operator like $\tops{badnat}{\kunit}{i}{a}{\tapp{\tvar{const}}{\tapp{\tvar{a}}{\tden{\titerm{-1}}{\ttype{Nat}{\tunit}}}}}$, breaking this invariant (and thus the guarantee that the recursor always terminates!)
-%To put it another way, if a collection of operators can be shown to be a full and faithful (that is, adequate) encoding of a particular type system, then it does not matter what other types there are in the system.
-
-We preclude such operations by a mechanism similar to the abstract type mechanism supported by ML-style module systems \cite{pfpl}.
+%%To put it another way, if a collection of operators can be shown to be a full and faithful (that is, adequate) encoding of a particular type system, then it does not matter what other types there are in the system.
+%
+%We preclude such operations by a mechanism similar to the abstract type mechanism supported by ML-style module systems \cite{pfpl}.
 By keeping a type family's representation schema private to the operators associated with it, they maintain full control over representation invariants.
@@
 Translations extracted from denotations via \textsf{case} analysis are tracked during this phase as \emph{abstract internal terms}, $\tvalof{\tau_1}{\tau_2}$ (see rule \textsc{dencase-eval-valid}).%
-
-Let us first review how expressions are assigned types and abstract translations.
+%
+%Let us first review how expressions are assigned types and abstract translations.
 Variables (\textsc{att-var}) translate directly to variables in the internal language, with the type determined by the typing context, $\Gamma$.
@@
 The actual representation of this type is only available from within operators in its family.
-
-Now we will consider the important operator application rule (\textsc{att-op}).
+%
+%Now we will consider the important operator application rule (\textsc{att-op}).
 This rule operates by extracting the appropriate operator definition, evaluating the operator index to a value, $\tauidx'$, and recursively assigning a type and abstract translation to each argument.
@@
 If this results in a valid denotation, compilation can proceed, but if an $\terr$ is produced, compilation will stop because no other rule will apply (in practice, we would display a type error at this point).
-
-Before a valid denotation produced in this way can be used, however, we check that it is \emph{representationally consistent}, meaning that the abstract translation, $\gabs$, is of an abstract internal type consistent with the abstraction representation of the type it is paired with, $\tau$.
+%
+%Before a valid denotation produced in this way can be used, however, we check that it is \emph{representationally consistent}, meaning that the abstract translation, $\gabs$, is of an abstract internal type consistent with the abstraction representation of the type it is paired with, $\tau$.
 The premise $\ddbar{\Xi_0,\fvar{Fam}}{\fvalCtx}{\trepof{\tau}}{\sabs}$ determines the abstract internal type and the premise $\checkRC{\gtCtx}{\Xi_0,\fvar{Fam}}{\fvalCtx}{\gabs}{\sabs}$ performs the internal type checking.
@@
 This could be exanded to support module-scoped visibility (we do not include modules in our calculus for simplicity), mutually-defined type families or type families that intentionally expose their representation schemas publicly (as $\fvar{Arrow}$ does).
-
-To clarify this fundamental mechanism let us examine how the expression $\eop{Nat}{s}{\tunit}{\eop{Nat}{z}{\tunit}{ }}$ will be processed during this phase.
+%
+%To clarify this fundamental mechanism let us examine how the expression $\eop{Nat}{s}{\tunit}{\eop{Nat}{z}{\tunit}{ }}$ will be processed during this phase.
 The inner application of $\opvar{z}$ will produce the abstract translation $0$ paired with the type $\ttype{Nat}{\tunit}$.
@@
 In contrast, if $\opvar{badnat}$ were used inside, the representation schema of $\fvar{Nat}$ would not be available, so we can only apply \textsc{hide-rep}, which does not give us enough information to give $-1$ a type.
-
-When the compiler next considers the outer application of $\opvar{s}$, it will pass the denotation $\tden{\titerm{0}}{\ttype{Nat}{\tunit}}$ into its definition.
+%
+%When the compiler next considers the outer application of $\opvar{s}$, it will pass the denotation $\tden{\titerm{0}}{\ttype{Nat}{\tunit}}$ into its definition.
 There, it binds the translation to the variable $\tvar{x}$ (within the $\tvar{pop\_final}$ helper function).
@@
 We encourage the reader to derive these judgements for the logic in the $\opvar{rec}$ operator to strengthen their understanding of this fundamental mechanism.
-
-We cannot leave abstract representations and abstract internal terms in the result of compilation, so a final deabstraction phase erases these, replacing them with their underlying internal types and terms in all contexts.
+%
+%We cannot leave abstract representations and abstract internal terms in the result of compilation, so a final deabstraction phase erases these, replacing them with their underlying internal types and terms in all contexts.
 Just as with abstract types in modules or existential types, there is no run-time overhead to this mechanism -- programs run at full speed.
 We give the deabstraction rules in the appendix due to their simplicity.
-
-\begin{figure}[t]
-\small
-%\vspace{-15pt}
-$\fbox{\inferrule{}{\tKindX{\tau}{\kappa}}}$
-\begin{mathpar}
-\small
-\inferrule[var-kind]{
-}{
-  \tKind{\tvarCtxX{t}{\kappa}}{\fvarCtx}{\tvar{t}}{\kappa}
-}
-
-\inferrule[k-arrow-intro]{
-  \tKind{\tvarCtxX{t}{\kappa_1}}{\fvarCtx}{\tau}{\kappa_2}
-}{
-  \tKindX{\tlam{t}{\kappa_1}{\tau}}{\karrow{\kappa_1}{\kappa_2}}
-}
-
-\inferrule[k-arrow-elim]{
-  \tKindX{\tau_1}{\karrow{\kappa_1}{\kappa_2}}\\
-  \tKindX{\tau_2}{\kappa_1}
-}{
-  \tKindX{\tapp{\tau_1}{\tau_2}}{\kappa_2}
-}
-\\
-\text{\color{gray} (standard statics for lists, integers, strings and products  omitted)}
-\\
+%
+%\begin{figure}[t]
+%\small
+%%\vspace{-15pt}
+%$\fbox{\inferrule{}{\tKindX{\tau}{\kappa}}}$
+%\begin{mathpar}
+%\small
+%\inferrule[var-kind]{
+%}{
+%  \tKind{\tvarCtxX{t}{\kappa}}{\fvarCtx}{\tvar{t}}{\kappa}
+%}
+%
+%\inferrule[k-arrow-intro]{
+%  \tKind{\tvarCtxX{t}{\kappa_1}}{\fvarCtx}{\tau}{\kappa_2}
+%}{
+%  \tKindX{\tlam{t}{\kappa_1}{\tau}}{\karrow{\kappa_1}{\kappa_2}}
+%}
+%
+%\inferrule[k-arrow-elim]{
+%  \tKindX{\tau_1}{\karrow{\kappa_1}{\kappa_2}}\\
+%  \tKindX{\tau_2}{\kappa_1}
+%}{
+%  \tKindX{\tapp{\tau_1}{\tau_2}}{\kappa_2}
+%}
+%\\
+%\text{\color{gray} (standard statics for lists, integers, strings and products  omitted)}
+%\\
+%%%
+%%%\inferrule{ }{
+%%%	\tKindX{\tstr{str}}{\kstr}
+%%%}(\text{str}^I_\tau)
+%%%
+%%%\inferrule{ }{
+%%%	\tKindX{\tunit}{\kunit}
+%%%}(\text{1}^I_\tau)
+%%%
+%%%\inferrule{
+%%%	\tKindX{\tau_1}{\kappa_1}\\
+%%%	\tKindX{\tau_2}{\kappa_2}
+%%%}{
+%%%	\tKindX{\tpair{\tau_1}{\tau_2}}{\kpair{\kappa_1}{\kappa_2}}
+%%%}({\times}^I_\tau)
+%%%
+%%%\inferrule{
+%%%	\tKindX{\tau}{\kpair}{\kappa_1}{\kappa_2}
+%%%}{
+%%%	\tKindX{\tfst{\tau}}{\kappa_1}
+%%%}({\times}^{E1}_\tau)
+%%%
+%%%\inferrule{
+%%%	\tKindX{\tau}{\kpair}{\kappa_1}{\kappa_2}
+%%%}{
+%%%	\tKindX{\tsnd{\tau}}{\kappa_2}
+%%%}({\times}^{E2}_\tau)
+%%%\inferrule{ }{
+%%%\inferrule{
+%%%	\tKindX{\tau_{1}}{\kappa_{1}}\\
+%%%	\tKindX{\tau_{2}}{\kappa_{2}}
+%%%}{
+%%%	\tKindX{\tpair{\tau_{1}}{\tau_{2}}}{\kpair{\kappa_{1}}{\kappa_{2}}}
+%%%}~(\times_\tau)
+%%%
+%%%\inferrule{
+%%%	\tKindX{\tau}{\kpair{\kappa_{1}}{\kappa_{2}}}
+%%%}{
+%%%	\tKindX{\tfst{\tau}}{\kappa_{1}}
+%%%}~(\text{fst}_\tau)
+%%%
+%%%\inferrule{
+%%%	\tKindX{\tau}{\kpair{\kappa_{1}}{\kappa_{2}}}
+%%%}{
+%%%	\tKindX{\tsnd{\tau}}{\kappa_{2}}
+%%%}~(\text{snd}_\tau)
+%%%
+%%% TODO: list
+%\inferrule[tl-equality]{
+%	\kEq{\kappa}\\
+%	\tKindX{\tau_1}{\kappa}\\
+%	\tKindX{\tau_2}{\kappa}\\\\
+%	\tKindX{\tau_3}{\kappa'}\\
+%	\tKindX{\tau_4}{\kappa'}
+%}{
+%	\tKindX{\tifeq{\tau_1}{\tau_2}{\kappa}{\tau_3}{\tau_4}}{\kappa'}
+%}
+%
+%\inferrule[type-intro]{
+%	\fvarOfTypeDf \in \fvarCtx\\\\
+%	\tKind{\tvarCtx}{\fvarCtx}{\tauidx}{\kappaidx}
+%}{
+%	\tKindX{\ttype{Fam}{\tauidx}}{\kTypeBlur}
+%}
+%
+%\inferrule[type-elim]{
+%	\fvarOfTypeDf \in \fvarCtx\\
+%	\tKindX{\tau}{\kTypeBlur}\\\\
+%	\tKind{\tvarCtxX{x}{\kappaidx}}{\fvarCtx}{\tau_0}{\kappa}\\
+%	\tKindX{\tau_1}{\kappa}
+%}{
+%	\tKind{\tvarCtx}{\fvarCtx}{\tfamcase{\tau}{Fam}{x}{\tau_0}{\tau_1}}{\kappa}
+%}
+%~~~~~~
+%\inferrule[den-intro-valid]{
+%	\tKindX{\tau_1}{\kITerm}\\\\
+%	\tKindX{\tau_2}{\kTypeBlur}
+%}{
+%	\tKindX{\tden{\tau_1}{\tau_2}}{\kDen}
+%}
+%
+%\inferrule[den-intro-err]{ }{
+%	\tKindX{\terr}{\kDen}
+%}
+%
+%\inferrule[den-elim]{
+%	\tKindX{\tau}{\kDen}\\\\
+%	\tKind{\tvarCtx, \tOfKind{\tvar{x}}{\kITerm}, \tOfKind{\tvar{t}}{\kTypeBlur}}{\fvarCtx}{\tau_1}{\kappa}\\
+%	\tKindX{\tau_2}{\kappa}
+%}{
+%	\tKindX{\tdencase{\tau}{x}{t}{\tau_1}{\tau_2}}{\kappa}
+%}
+%
+%\inferrule[iterm-intro]{
+%	\isIterm{\tvarCtx}{\emptyctx}{\fvarCtx}{\gamma}
+%}{
+%	\tKindX{\titerm{\gamma}}{\kITerm}
+%}
+%
+%\inferrule[itype intro]{
+%	\isItype{\tvarCtx}{\fvarCtx}{\sigma}
+%}{
+%	\tKindX{\titype{\sigma}}{\kIType}
+%}
+%\end{mathpar}
+%$\fbox{$\kEq{\kappa}$}$
+%\begin{mathpar}
+%\inferrule[t-eq]{ }{
+%	\kEq{\kTypeBlur}
+%}
+%
+%\inferrule[z-eq]{ }{
+%	\kEq{\dint}
+%}
+%
+%\inferrule[s-eq]{ }{
+%	\kEq{\kstr}
+%}
+%
+%\inferrule[u-eq]{ }{
+%	\kEq{\kunit}
+%}
+%
+%\inferrule[p-eq]{
+%	\kEq{\kappa_1}\\\\
+%	\kEq{\kappa_2}
+%}{
+%	\kEq{\kpair{\kappa_1}{\kappa_2}}
+%}
+%
+%\inferrule[l-eq]{
+%	\kEq{\kappa}
+%}{
+%	\kEq{\klist{\kappa}}
+%}
+%\end{mathpar}
+%$\fbox{$\isItermX{\gamma}$}$
+%\begin{mathpar}
+%\inferrule[i-var-kinding]{ }{
+%	\isIterm{\tvarCtx}{\eivarCtxX{x}}{\fvarCtx}{\evar{x}}
+%}
+%
+%\inferrule[i-lam-kinding]{
+%	\isItypeX{\sigma}\\\\
+%	\isIterm{\tvarCtx}{\eivarCtxX{x}}{\fvarCtx}{\gamma}
+%}{
+%	\isItermX{\ilam{\evar{x}}{\sigma}{\gamma}}
+%}
+%
+%\inferrule[i-fix-kinding]{
+%	\isItypeX{\sigma}\\\\
+%	\isIterm{\tvarCtx}{\eivarCtxX{x}}{\fvarCtx}{\gamma}
+%}{
+%	\isItermX{\ifix{\evar{f}}{\sigma}{\gamma}}
+%}
+%\\
+%\text{\color{gray} (omitted forms have trivially recursive rules)}\\\vspace{-3pt}
+%\\
+%\inferrule[iterm-dereify-kinding]{
+%	\tKindX{\tau}{\kITerm}
+%}{
+%	\isItermX{\iup{\tau}}
+%}
+%
+%\inferrule[abs-trans-kinding]{
+%	\tKindX{\tauiterm}{\kITerm}\\
+%	\tKindX{\tautype}{\kTypeBlur}
+%}{
+%	\isItermX{\tvalof{\tauiterm}{\tautype}}
+%}
+%\end{mathpar}
+%$\fbox{$\isItypeX{\sigma}$}$
+%\begin{mathpar}
+%\inferrule[i-int-kinding]{ }{
+%	\isItypeX{\dint}
+%}
+%
+%\inferrule[i-prod-kinding]{
+%	\isItypeX{\sigma_1}\\\\
+%	\isItypeX{\sigma_2}
+%}{
+%	\isItypeX{\dpair{\sigma_1}{\sigma_2}}
+%}
+%
+%\inferrule[i-arrow-kinding]{
+%	\isItypeX{\sigma_1}\\\\
+%	\isItypeX{\sigma_2}
+%}{
+%	\isItypeX{\darrow{\sigma_1}{\sigma_2}}
+%}
+%
+%\inferrule[itype-dereify-kinding]{
+%	\tKindX{\tau}{\kIType}
+%}{
+%	\isItypeX{\dup{\tau}}
+%}
+%
+%\inferrule[abs-rep-kinding]{
+%	\tKindX{\tau}{\kTypeBlur}
+%}{
+%	\isItypeX{\trepof{\tau}}
+%}
+%\end{mathpar}
+%\caption{\small Kinding for type-level terms}
+%\label{tlkind}
+%\end{figure}
+%
+%\begin{figure}
+%\small
+%$\fbox{$\tEvalX{\tau}{\tau'}$}$
+%\begin{mathpar}
+%\inferrule[tl-lam-eval]{ }{
+%	\tEvalX{\tlam{t}{\kappa}{\tau}}{\tlam{t}{\kappa}{\tau}}
+%}
+%
+%\inferrule[tl-ap-eval]{
+%	\tEvalX{\tau_1}{\tlam{t}{\kappa}{\tau}}\\
+%	\tEvalX{\tau_2}{\tau_2'}\\
+%	\tEvalX{\subst{\tau_2'}{\tvar{t}}{\tau}}{\tau'}
+%}{
+%	\tEvalX{\tapp{\tau_1}{\tau_2}}{\tau'}
+%}
+%
+%\text{\color{gray} (standard evaluation rules for integers, strings, products and lists omitted)}
+%
+%\inferrule[tl-eq-eval-equal]{
+%	\tEvalX{\tau_1}{\tau_1'}\\
+%	\tEvalX{\tau_2}{\tau_1'}\\
+%	\tEvalX{\tau_3}{\tau_3'}
+%}{
+%	\tEvalX{\tifeq{\tau_1}{\tau_2}{\kappa}{\tau_3}{\tau_4}}{\tau_3'}
+%}
+%
+%\inferrule[tl-eq-eval-inequal]{
+%	\tEvalX{\tau_1}{\tau_1'}\\
+%	\tEvalX{\tau_2}{\tau_2'}\\
+%	\tau_1' \neq \tau_2'\\
+%	\tEvalX{\tau_4}{\tau_4'}
+%}{
+%	\tEvalX{\tifeq{\tau_1}{\tau_2}{\kappa}{\tau_3}{\tau_4}}{\tau_4'}
+%}
+%
+%\inferrule[type-eval]{
+%	\tEvalX{\tauidx}{\tauidx'}
+%}{
+%	\tEvalX{\ttype{Fam}{\tauidx}}{\ttype{Fam}{\tauidx'}}
+%}
+%
+%\inferrule[famcase-eval-match]{
+%	\tEvalX{\tau}{\ttype{Fam}{\tauidx}}\\
+%	\tEvalX{\subst{\tauidx}{\tvar{x}}{\tau_1}}{\tau_1'}
+%}{
+%	\tEvalX{\tfamcase{\tau}{Fam}{x}{\tau_1}{\tau_2}}{\tau_1'}
+%}
+%
+%\inferrule[famcase-eval-fail]{
+%	\tEvalX{\tau}{\ttype{Fam'}{\tauidx}}\\
+%	\fvar{Fam} \neq \fvar{Fam'}\\
+%	\tEvalX{\tau_2}{\tau_2'}
+%}{
+%	\tEvalX{\tfamcase{\tau}{Fam}{x}{\tau_1}{\tau_2}}{\tau_2'}
+%}
+%
+%\inferrule[den-valid-eval]{
+%	\tEvalX{\tau_1}{\tau_1'}\\
+%	\tEvalX{\tau_2}{\tau_2'}
+%}{
+%	\tEvalX{\tden{\tau_1}{\tau_2}}{\tden{\tau_1'}{\tau_2'}}
+%}
+%
+%\inferrule[den-err-eval]{ }{
+%	\tEvalX{\terr}{\terr}
+%}
+%
+%\inferrule[dencase-eval-valid]{
+%	\tEvalX{\tau}{\tden{\tauiterm}{\tautype}}\\\\
+%	\tEvalX{[\titerm{\tvalof{\tauiterm}{\tautype}}/\tvar{x}, \tautype/\tvar{t}]\tau_1}{\tau_1'}
+%}{
+%	\tEvalX{\tdencase{\tau}{x}{t}{\tau_1}{\tau_2}}{\tau_1'}
+%}
+%
+%\inferrule[dencase-eval-err]{
+%	\tEvalX{\tau}{\terr}\\
+%	\tEvalX{\tau_2}{\tau_2'}
+%}{
+%	\tEvalX{\tdencase{\tauden}{y}{x}{\tau_1}{\tau_2}}{\tau_2'}
+%}
+%
+%\inferrule[iterm-reify]{
+%	\tiEvalX{\gamma}{\gamma'}
+%}{
+%	\tEvalX{\titerm{\gamma}}{\titerm{\gamma'}}
+%}
+%
+%\inferrule[itype-reify]{
+%	\tiEvalX{\sigma}{\sigma'}
+%}{
+%	\tEvalX{\titype{\sigma}}{\titype{\sigma'}}
+%}
+%\end{mathpar}
+%$\fbox{$\tiEvalX{\gamma}{\gamma'}$}$
+%\begin{mathpar}
+%\inferrule[i-var-eval]{ }{
+%	\tiEvalX{\evar{x}}{\evar{x}}
+%}
+%
+%\inferrule[i-lam-eval]{
+%	\tiEvalX{\sigma}{\sigma'}\\
+%	\tiEvalX{\gamma}{\gamma'}
+%}{
+%	\tiEvalX{\ilam{\evar{x}}{\sigma}{\gamma}}{\ilam{\evar{x}}{\sigma'}{\gamma'}}
+%}
+%
+%\inferrule[i-fix-eval]{
+%	\tiEvalX{\sigma}{\sigma'}\\
+%	\tiEvalX{\gamma}{\gamma'}
+%}{
+%	\tiEvalX{\ifix{\evar{f}}{\sigma}{\gamma}}{\ifix{\evar{f}}{\sigma'}{\gamma'}}
+%}
+%\\
+%\text{\color{gray} (omitted forms have trivially recursive rules)}
+%\\
+%\inferrule[iterm unquote eval]{
+%	\tEvalX{\tau}{\tau'}
+%}{
+%	\tiEvalX{\iup{\tau}}{\iup{\tau'}}
+%}
+%
+%\inferrule[val from den eval]{
+%	\tEvalX{\tauiterm}{\tauiterm'}\\
+%	\tEvalX{\tautype}{\tautype'}
+%}{
+%	\tiEvalX{\tvalof{\tauiterm}{\tautype}}{\tvalof{\tauiterm'}{\tautype'}}
+%}
+%\end{mathpar}
+%$\fbox{$\tiEvalX{\sigma}{\sigma'}$}$
+%\begin{mathpar}
+%\inferrule[i-int-eval]{ }{
+%	\tiEvalX{\dint}{\dint}
+%}
+%
+%\inferrule[i-prod-eval]{
+%	\tiEvalX{\sigma_1}{\sigma_1'}\\
+%	\tiEvalX{\sigma_2}{\sigma_2'}
+%}{
+%	\tiEvalX{\dpair{\sigma_1}{\sigma_2}}{\dpair{\sigma_1'}{\sigma_2'}}
+%}
+%
+%\inferrule[i-arrow-eval]{
+%	\tiEvalX{\sigma_1}{\sigma_1'}\\
+%	\tiEvalX{\sigma_2}{\sigma_2'}
+%}{
+%	\tiEvalX{\darrow{\sigma_1}{\sigma_2}}{\darrow{\sigma_1'}{\sigma_2'}}
+%}
+%\\
+%\inferrule[itype-eval]{
+%	\tEvalX{\tau}{\tau'}
+%}{
+%	\tiEvalX{\dup{\tau}}{\dup{\tau'}}
+%}
+%
+%\inferrule[abs-rep-eval]{
+%	\tEvalX{\tau}{\tau'}
+%}{
+%	\tiEvalX{\trepof{\tau}}{\trepof{\tau'}}
+%}
+%\end{mathpar}
+%\caption{\small Evaluation semantics for type-level terms}
+%\label{tleval}
+%\end{figure}
+%
+%\begin{figure}[t]
+%\small
+%$\fbox{\inferrule{}{\concrep{\fvalCtx}{\tau}{\sigma}}}$
+%\begin{mathpar}
+%\inferrule[get-rep]{
+%	\fval{Fam}{\theta}{i}{\tau} \in \fvalCtx\\
+%	\tEvalX{[\tauidx/\tvar{i}]\tau}{\titype{\sigma}}\\
+%	\ddbarX{\sigma}{\sconc}
+%}{
+%	\concrep{\fvalCtx}{\ttype{Fam}{\tauidx}}{\sconc}
+%}
+%\end{mathpar}
+%$\fbox{\inferrule{}{\ddbarX{\sigma}{\sigma'}}}$
+%~~~$\fvalCtx ::= \fvalCtx_0 \pipe \fvalCtx, \fvalDf$~~~
+%~~~$\Xi ::= \Xi_0 \pipe \Xi, \fvar{Fam}$~~~
+%~~~$\Xi_0 := \fvar{Arrow}$
+%\begin{mathpar}
+%\inferrule[abs-int]{ }{
+%	\ddbarX{\dint}{\dint}
+%}
+%
+%\inferrule[abs-arrow]{
+%	\ddbarX{\sigma_1}{\sigma_1'}\\
+%	\ddbarX{\sigma_2}{\sigma_2'}
+%}{
+%	\ddbarX{\darrow{\sigma_1}{\sigma_2}}{\darrow{\sigma_1'}{\sigma_2'}}
+%}
+%
+%\inferrule[abs-prod]{
+%	\ddbarX{\sigma_1}{\sigma_1'}\\
+%	\ddbarX{\sigma_2}{\sigma_2'}
+%}{
+%	\ddbarX{\dpair{\sigma_1}{\sigma_2}}{\dpair{\sigma_1'}{\sigma_2'}}
+%}
+%
+%\inferrule[itype-inverse]{
+%	\ddbarX{\sigma}{\sabs}
+%}{
+%	\ddbarX{\dup{\titype{\sigma}}}{\sabs}
+%}
+%
+%\inferrule[show-rep]{
+%	\fvar{Fam} \in \Xi\\
+%	\concrep{\fvalCtx}{\ttype{Fam}{\tauidx}}{\sconc}
+%}{
+%	\ddbarX{\trepof{\ttype{Fam}{\tauidx}}}{\sconc}
+%}
+%
+%\inferrule[hide-rep]{
+%	\fvar{Fam} \notin \Xi
+%}{
+%	\ddbarX{\trepof{\ttype{Fam}{\tauidx}}}{\trepof{\ttype{Fam}{\tauidx}}}
+%}
+%\end{mathpar}
+%$\fbox{\inferrule{}{\eCtxTogCtxX{\etCtx}{\gtCtx}}}$
+%~~~$\gtCtx ::= \emptyctx \pipe \gtCtxX{x}{\sigma}$
+%\begin{mathpar}
+%\inferrule[abs-empty]{ }{
+%	\eCtxTogCtxX{\emptyctx}{\emptyctx}
+%}
+%
+%\inferrule[abs-ctx]{
+%	\eCtxTogCtxX{\etCtx}{\gtCtx}\\
+%	\ddbarX{\trepof{\tau}}{\sigma}
+%}{
+%	\eCtxTogCtxX{\etCtxX{x}{\tau}}{\gtCtxX{x}{\sigma}}
+%}
+%\end{mathpar}
+%$\fbox{\inferrule{}{\checkRCX{\gamma}{\sigma}}}$
+%\begin{mathpar}
+%\inferrule[abs-i-var]{ }{
+%	\checkRC{\gtCtxX{x}{\sigma}}{\Xi}{\fvalCtx}{\evar{x}}{\sigma}
+%}
+%
+%\inferrule[abs-i-lam]{
+%	\ddbarX{\sigma_1}{\sigma_1'}\\
+%	\checkRC{\gtCtxX{x}{\sigma_1'}}{\Xi}{\fvalCtx}{\gamma}{\sigma_2}
+%}{
+%	\checkRCX{\ilam{x}{\sigma_1}{\gamma}}{\darrow{\sigma_1'}{\sigma_2}}
+%}
+%
+%\inferrule[abs-i-ap]{
+%	\checkRCX{\gamma_1}{\darrow{\sigma_1}{\sigma_2}}\\
+%	\checkRCX{\gamma_2}{\sigma_1}
+%}{
+%	\checkRCX{\iapp{\gamma_1}{\gamma_2}}{\sigma_2}
+%}
+%
+%\inferrule[abs-i-fix]{
+%	\ddbarX{\sigma}{\sigma'}\\
+%	\checkRC{\gtCtxX{x}{\sigma'}}{\Xi}{\fvalCtx}{\gamma}{\sigma'}
+%}{
+%	\checkRCX{\ifix{x}{\sigma}{\gabs}}{\sigma'}
+%}
+%\\
+%\text{\color{gray} (standard statics for integers and products omitted)}
+%\\
+%%\inferrule[abs int]{ }{
+%%	\checkRCX{\iintlit}{\dint}
+%%}
 %%
-%%\inferrule{ }{
-%%	\tKindX{\tstr{str}}{\kstr}
-%%}(\text{str}^I_\tau)
+%%\inferrule[abs op]{
+%%	\checkRCX{\gamma_1}{\dint}\\
+%%	\checkRCX{\gamma_2}{\dint}
+%%}{
+%%	\checkRCX{\iop{\gamma_1}{\gamma_2}}{\dint}
+%%}
 %%
-%%\inferrule{ }{
-%%	\tKindX{\tunit}{\kunit}
-%%}(\text{1}^I_\tau)
+%%\inferrule[abs pair]{
+%%	\checkRCX{\gamma_1}{\sigma_1}\\
+%%	\checkRCX{\gamma_2}{\sigma_2}
+%%}{
+%%	\checkRCX{\ipair{\gamma_1}{\gamma_2}}{\dpair{\sigma_1}{\sigma_2}}
+%%}
 %%
-%%\inferrule{
-%%	\tKindX{\tau_1}{\kappa_1}\\
-%%	\tKindX{\tau_2}{\kappa_2}
+%%\inferrule[abs fst]{
+%%	\checkRCX{\gamma}{\dpair{\sigma_1}{\sigma_2}}
 %%}{
-%%	\tKindX{\tpair{\tau_1}{\tau_2}}{\kpair{\kappa_1}{\kappa_2}}
-%%}({\times}^I_\tau)
+%%	\checkRCX{\ifst{\gamma}}{\sigma_1}
+%%}
 %%
-%%\inferrule{
-%%	\tKindX{\tau}{\kpair}{\kappa_1}{\kappa_2}
+%%\inferrule[abs snd]{
+%%	\checkRCX{\gamma}{\dpair{\sigma_1}{\sigma_2}}
 %%}{
-%%	\tKindX{\tfst{\tau}}{\kappa_1}
-%%}({\times}^{E1}_\tau)
+%%	\checkRCX{\isnd{\gamma}}{\sigma_2}
+%%}
 %%
-%%\inferrule{
-%%	\tKindX{\tau}{\kpair}{\kappa_1}{\kappa_2}
-%%}{
-%%	\tKindX{\tsnd{\tau}}{\kappa_2}
-%%}({\times}^{E2}_\tau)
-%%\inferrule{ }{
-%%\inferrule{
-%%	\tKindX{\tau_{1}}{\kappa_{1}}\\
-%%	\tKindX{\tau_{2}}{\kappa_{2}}
-%%}{
-%%	\tKindX{\tpair{\tau_{1}}{\tau_{2}}}{\kpair{\kappa_{1}}{\kappa_{2}}}
-%%}~(\times_\tau)
+%% \inferrule[abs-int-eq]{
+%% 	\checkRCX{\gamma_1}{\dint}\\
+%% 	\checkRCX{\gamma_2}{\dint}\\\\
+%% 	\checkRCX{\gamma_3}{\sigma}\\
+%% 	\checkRCX{\gamma_4}{\sigma}
+%% }{
+%% 	\checkRCX{\iIfEq{\gamma_1}{\gamma_2}{\dint}{\gamma_3}{\gamma_4}}{\sigma}
+%% }
+%\inferrule[abs-iterm-inverse]{
+%	\checkRCX{\gamma}{\sigma}
+%}{
+%	\checkRCX{\iup{\titerm{\gamma}}}{\sigma}
+%}
+%
+%\inferrule[show-trans]{
+%	\fvar{Fam} \in \Xi\\
+%	\concrep{\fvalCtx}{\ttype{Fam}{\tauidx}}{\sigma}\\
+%	\checkRCX{\gamma}{\sigma}
+%}{
+%	\checkRCX{\tvalof{\titerm{\gamma}}{\ttype{Fam}{\tauidx}}}{\sigma}
+%}
+%
+%\inferrule[hide-trans]{
+%	\fvar{Fam} \notin \Xi\\
+%	\concrep{\fvalCtx}{\ttype{Fam}{\tauidx}}{\sigma}\\
+%	\checkRCX{\gamma}{\sigma}
+%}{
+%	\checkRCX{\tvalof{\titerm{\gamma}}{\ttype{Fam}{\tauidx}}}{\trepof{\ttype{Fam}{\tauidx}}}
+%}
+%\end{mathpar}
+%\caption{\small Abstracted internal typing}
+%\label{ait}
+%\end{figure}
 %%
-%%\inferrule{
-%%	\tKindX{\tau}{\kpair{\kappa_{1}}{\kappa_{2}}}
-%%}{
-%%	\tKindX{\tfst{\tau}}{\kappa_{1}}
-%%}~(\text{fst}_\tau)
-%%
-%%\inferrule{
-%%	\tKindX{\tau}{\kpair{\kappa_{1}}{\kappa_{2}}}
-%%}{
-%%	\tKindX{\tsnd{\tau}}{\kappa_{2}}
-%%}~(\text{snd}_\tau)
-%%
-%% TODO: list
-\inferrule[tl-equality]{
-	\kEq{\kappa}\\
-	\tKindX{\tau_1}{\kappa}\\
-	\tKindX{\tau_2}{\kappa}\\\\
-	\tKindX{\tau_3}{\kappa'}\\
-	\tKindX{\tau_4}{\kappa'}
-}{
-	\tKindX{\tifeq{\tau_1}{\tau_2}{\kappa}{\tau_3}{\tau_4}}{\kappa'}
-}
-
-\inferrule[type-intro]{
-	\fvarOfTypeDf \in \fvarCtx\\\\
-	\tKind{\tvarCtx}{\fvarCtx}{\tauidx}{\kappaidx}
-}{
-	\tKindX{\ttype{Fam}{\tauidx}}{\kTypeBlur}
-}
-
-\inferrule[type-elim]{
-	\fvarOfTypeDf \in \fvarCtx\\
-	\tKindX{\tau}{\kTypeBlur}\\\\
-	\tKind{\tvarCtxX{x}{\kappaidx}}{\fvarCtx}{\tau_0}{\kappa}\\
-	\tKindX{\tau_1}{\kappa}
-}{
-	\tKind{\tvarCtx}{\fvarCtx}{\tfamcase{\tau}{Fam}{x}{\tau_0}{\tau_1}}{\kappa}
-}
-~~~~~~
-\inferrule[den-intro-valid]{
-	\tKindX{\tau_1}{\kITerm}\\\\
-	\tKindX{\tau_2}{\kTypeBlur}
-}{
-	\tKindX{\tden{\tau_1}{\tau_2}}{\kDen}
-}
-
-\inferrule[den-intro-err]{ }{
-	\tKindX{\terr}{\kDen}
-}
-
-\inferrule[den-elim]{
-	\tKindX{\tau}{\kDen}\\\\
-	\tKind{\tvarCtx, \tOfKind{\tvar{x}}{\kITerm}, \tOfKind{\tvar{t}}{\kTypeBlur}}{\fvarCtx}{\tau_1}{\kappa}\\
-	\tKindX{\tau_2}{\kappa}
-}{
-	\tKindX{\tdencase{\tau}{x}{t}{\tau_1}{\tau_2}}{\kappa}
-}
-
-\inferrule[iterm-intro]{
-	\isIterm{\tvarCtx}{\emptyctx}{\fvarCtx}{\gamma}
-}{
-	\tKindX{\titerm{\gamma}}{\kITerm}
-}
-
-\inferrule[itype intro]{
-	\isItype{\tvarCtx}{\fvarCtx}{\sigma}
-}{
-	\tKindX{\titype{\sigma}}{\kIType}
-}
-\end{mathpar}
-$\fbox{$\kEq{\kappa}$}$
-\begin{mathpar}
-\inferrule[t-eq]{ }{
-	\kEq{\kTypeBlur}
-}
-
-\inferrule[z-eq]{ }{
-	\kEq{\dint}
-}
-
-\inferrule[s-eq]{ }{
-	\kEq{\kstr}
-}
-
-\inferrule[u-eq]{ }{
-	\kEq{\kunit}
-}
-
-\inferrule[p-eq]{
-	\kEq{\kappa_1}\\\\
-	\kEq{\kappa_2}
-}{
-	\kEq{\kpair{\kappa_1}{\kappa_2}}
-}
-
-\inferrule[l-eq]{
-	\kEq{\kappa}
-}{
-	\kEq{\klist{\kappa}}
-}
-\end{mathpar}
-$\fbox{$\isItermX{\gamma}$}$
-\begin{mathpar}
-\inferrule[i-var-kinding]{ }{
-	\isIterm{\tvarCtx}{\eivarCtxX{x}}{\fvarCtx}{\evar{x}}
-}
-
-\inferrule[i-lam-kinding]{
-	\isItypeX{\sigma}\\\\
-	\isIterm{\tvarCtx}{\eivarCtxX{x}}{\fvarCtx}{\gamma}
-}{
-	\isItermX{\ilam{\evar{x}}{\sigma}{\gamma}}
-}
-
-\inferrule[i-fix-kinding]{
-	\isItypeX{\sigma}\\\\
-	\isIterm{\tvarCtx}{\eivarCtxX{x}}{\fvarCtx}{\gamma}
-}{
-	\isItermX{\ifix{\evar{f}}{\sigma}{\gamma}}
-}
-\\
-\text{\color{gray} (omitted forms have trivially recursive rules)}\\\vspace{-3pt}
-\\
-\inferrule[iterm-dereify-kinding]{
-	\tKindX{\tau}{\kITerm}
-}{
-	\isItermX{\iup{\tau}}
-}
-
-\inferrule[abs-trans-kinding]{
-	\tKindX{\tauiterm}{\kITerm}\\
-	\tKindX{\tautype}{\kTypeBlur}
-}{
-	\isItermX{\tvalof{\tauiterm}{\tautype}}
-}
-\end{mathpar}
-$\fbox{$\isItypeX{\sigma}$}$
-\begin{mathpar}
-\inferrule[i-int-kinding]{ }{
-	\isItypeX{\dint}
-}
-
-\inferrule[i-prod-kinding]{
-	\isItypeX{\sigma_1}\\\\
-	\isItypeX{\sigma_2}
-}{
-	\isItypeX{\dpair{\sigma_1}{\sigma_2}}
-}
-
-\inferrule[i-arrow-kinding]{
-	\isItypeX{\sigma_1}\\\\
-	\isItypeX{\sigma_2}
-}{
-	\isItypeX{\darrow{\sigma_1}{\sigma_2}}
-}
-
-\inferrule[itype-dereify-kinding]{
-	\tKindX{\tau}{\kIType}
-}{
-	\isItypeX{\dup{\tau}}
-}
-
-\inferrule[abs-rep-kinding]{
-	\tKindX{\tau}{\kTypeBlur}
-}{
-	\isItypeX{\trepof{\tau}}
-}
-\end{mathpar}
-\caption{\small Kinding for type-level terms}
-\label{tlkind}
-\end{figure}
-
-\begin{figure}
-\small
-$\fbox{$\tEvalX{\tau}{\tau'}$}$
-\begin{mathpar}
-\inferrule[tl-lam-eval]{ }{
-	\tEvalX{\tlam{t}{\kappa}{\tau}}{\tlam{t}{\kappa}{\tau}}
-}
-
-\inferrule[tl-ap-eval]{
-	\tEvalX{\tau_1}{\tlam{t}{\kappa}{\tau}}\\
-	\tEvalX{\tau_2}{\tau_2'}\\
-	\tEvalX{\subst{\tau_2'}{\tvar{t}}{\tau}}{\tau'}
-}{
-	\tEvalX{\tapp{\tau_1}{\tau_2}}{\tau'}
-}
-
-\text{\color{gray} (standard evaluation rules for integers, strings, products and lists omitted)}
-
-\inferrule[tl-eq-eval-equal]{
-	\tEvalX{\tau_1}{\tau_1'}\\
-	\tEvalX{\tau_2}{\tau_1'}\\
-	\tEvalX{\tau_3}{\tau_3'}
-}{
-	\tEvalX{\tifeq{\tau_1}{\tau_2}{\kappa}{\tau_3}{\tau_4}}{\tau_3'}
-}
-
-\inferrule[tl-eq-eval-inequal]{
-	\tEvalX{\tau_1}{\tau_1'}\\
-	\tEvalX{\tau_2}{\tau_2'}\\
-	\tau_1' \neq \tau_2'\\
-	\tEvalX{\tau_4}{\tau_4'}
-}{
-	\tEvalX{\tifeq{\tau_1}{\tau_2}{\kappa}{\tau_3}{\tau_4}}{\tau_4'}
-}
-
-\inferrule[type-eval]{
-	\tEvalX{\tauidx}{\tauidx'}
-}{
-	\tEvalX{\ttype{Fam}{\tauidx}}{\ttype{Fam}{\tauidx'}}
-}
-
-\inferrule[famcase-eval-match]{
-	\tEvalX{\tau}{\ttype{Fam}{\tauidx}}\\
-	\tEvalX{\subst{\tauidx}{\tvar{x}}{\tau_1}}{\tau_1'}
-}{
-	\tEvalX{\tfamcase{\tau}{Fam}{x}{\tau_1}{\tau_2}}{\tau_1'}
-}
-
-\inferrule[famcase-eval-fail]{
-	\tEvalX{\tau}{\ttype{Fam'}{\tauidx}}\\
-	\fvar{Fam} \neq \fvar{Fam'}\\
-	\tEvalX{\tau_2}{\tau_2'}
-}{
-	\tEvalX{\tfamcase{\tau}{Fam}{x}{\tau_1}{\tau_2}}{\tau_2'}
-}
-
-\inferrule[den-valid-eval]{
-	\tEvalX{\tau_1}{\tau_1'}\\
-	\tEvalX{\tau_2}{\tau_2'}
-}{
-	\tEvalX{\tden{\tau_1}{\tau_2}}{\tden{\tau_1'}{\tau_2'}}
-}
-
-\inferrule[den-err-eval]{ }{
-	\tEvalX{\terr}{\terr}
-}
-
-\inferrule[dencase-eval-valid]{
-	\tEvalX{\tau}{\tden{\tauiterm}{\tautype}}\\\\
-	\tEvalX{[\titerm{\tvalof{\tauiterm}{\tautype}}/\tvar{x}, \tautype/\tvar{t}]\tau_1}{\tau_1'}
-}{
-	\tEvalX{\tdencase{\tau}{x}{t}{\tau_1}{\tau_2}}{\tau_1'}
-}
-
-\inferrule[dencase-eval-err]{
-	\tEvalX{\tau}{\terr}\\
-	\tEvalX{\tau_2}{\tau_2'}
-}{
-	\tEvalX{\tdencase{\tauden}{y}{x}{\tau_1}{\tau_2}}{\tau_2'}
-}
-
-\inferrule[iterm-reify]{
-	\tiEvalX{\gamma}{\gamma'}
-}{
-	\tEvalX{\titerm{\gamma}}{\titerm{\gamma'}}
-}
-
-\inferrule[itype-reify]{
-	\tiEvalX{\sigma}{\sigma'}
-}{
-	\tEvalX{\titype{\sigma}}{\titype{\sigma'}}
-}
-\end{mathpar}
-$\fbox{$\tiEvalX{\gamma}{\gamma'}$}$
-\begin{mathpar}
-\inferrule[i-var-eval]{ }{
-	\tiEvalX{\evar{x}}{\evar{x}}
-}
-
-\inferrule[i-lam-eval]{
-	\tiEvalX{\sigma}{\sigma'}\\
-	\tiEvalX{\gamma}{\gamma'}
-}{
-	\tiEvalX{\ilam{\evar{x}}{\sigma}{\gamma}}{\ilam{\evar{x}}{\sigma'}{\gamma'}}
-}
-
-\inferrule[i-fix-eval]{
-	\tiEvalX{\sigma}{\sigma'}\\
-	\tiEvalX{\gamma}{\gamma'}
-}{
-	\tiEvalX{\ifix{\evar{f}}{\sigma}{\gamma}}{\ifix{\evar{f}}{\sigma'}{\gamma'}}
-}
-\\
-\text{\color{gray} (omitted forms have trivially recursive rules)}
-\\
-\inferrule[iterm unquote eval]{
-	\tEvalX{\tau}{\tau'}
-}{
-	\tiEvalX{\iup{\tau}}{\iup{\tau'}}
-}
-
-\inferrule[val from den eval]{
-	\tEvalX{\tauiterm}{\tauiterm'}\\
-	\tEvalX{\tautype}{\tautype'}
-}{
-	\tiEvalX{\tvalof{\tauiterm}{\tautype}}{\tvalof{\tauiterm'}{\tautype'}}
-}
-\end{mathpar}
-$\fbox{$\tiEvalX{\sigma}{\sigma'}$}$
-\begin{mathpar}
-\inferrule[i-int-eval]{ }{
-	\tiEvalX{\dint}{\dint}
-}
-
-\inferrule[i-prod-eval]{
-	\tiEvalX{\sigma_1}{\sigma_1'}\\
-	\tiEvalX{\sigma_2}{\sigma_2'}
-}{
-	\tiEvalX{\dpair{\sigma_1}{\sigma_2}}{\dpair{\sigma_1'}{\sigma_2'}}
-}
-
-\inferrule[i-arrow-eval]{
-	\tiEvalX{\sigma_1}{\sigma_1'}\\
-	\tiEvalX{\sigma_2}{\sigma_2'}
-}{
-	\tiEvalX{\darrow{\sigma_1}{\sigma_2}}{\darrow{\sigma_1'}{\sigma_2'}}
-}
-\\
-\inferrule[itype-eval]{
-	\tEvalX{\tau}{\tau'}
-}{
-	\tiEvalX{\dup{\tau}}{\dup{\tau'}}
-}
-
-\inferrule[abs-rep-eval]{
-	\tEvalX{\tau}{\tau'}
-}{
-	\tiEvalX{\trepof{\tau}}{\trepof{\tau'}}
-}
-\end{mathpar}
-\caption{\small Evaluation semantics for type-level terms}
-\label{tleval}
-\end{figure}
-
-\begin{figure}[t]
-\small
-$\fbox{\inferrule{}{\concrep{\fvalCtx}{\tau}{\sigma}}}$
-\begin{mathpar}
-\inferrule[get-rep]{
-	\fval{Fam}{\theta}{i}{\tau} \in \fvalCtx\\
-	\tEvalX{[\tauidx/\tvar{i}]\tau}{\titype{\sigma}}\\
-	\ddbarX{\sigma}{\sconc}
-}{
-	\concrep{\fvalCtx}{\ttype{Fam}{\tauidx}}{\sconc}
-}
-\end{mathpar}
-$\fbox{\inferrule{}{\ddbarX{\sigma}{\sigma'}}}$
-~~~$\fvalCtx ::= \fvalCtx_0 \pipe \fvalCtx, \fvalDf$~~~
-~~~$\Xi ::= \Xi_0 \pipe \Xi, \fvar{Fam}$~~~
-~~~$\Xi_0 := \fvar{Arrow}$
-\begin{mathpar}
-\inferrule[abs-int]{ }{
-	\ddbarX{\dint}{\dint}
-}
-
-\inferrule[abs-arrow]{
-	\ddbarX{\sigma_1}{\sigma_1'}\\
-	\ddbarX{\sigma_2}{\sigma_2'}
-}{
-	\ddbarX{\darrow{\sigma_1}{\sigma_2}}{\darrow{\sigma_1'}{\sigma_2'}}
-}
-
-\inferrule[abs-prod]{
-	\ddbarX{\sigma_1}{\sigma_1'}\\
-	\ddbarX{\sigma_2}{\sigma_2'}
-}{
-	\ddbarX{\dpair{\sigma_1}{\sigma_2}}{\dpair{\sigma_1'}{\sigma_2'}}
-}
-
-\inferrule[itype-inverse]{
-	\ddbarX{\sigma}{\sabs}
-}{
-	\ddbarX{\dup{\titype{\sigma}}}{\sabs}
-}
-
-\inferrule[show-rep]{
-	\fvar{Fam} \in \Xi\\
-	\concrep{\fvalCtx}{\ttype{Fam}{\tauidx}}{\sconc}
-}{
-	\ddbarX{\trepof{\ttype{Fam}{\tauidx}}}{\sconc}
-}
-
-\inferrule[hide-rep]{
-	\fvar{Fam} \notin \Xi
-}{
-	\ddbarX{\trepof{\ttype{Fam}{\tauidx}}}{\trepof{\ttype{Fam}{\tauidx}}}
-}
-\end{mathpar}
-$\fbox{\inferrule{}{\eCtxTogCtxX{\etCtx}{\gtCtx}}}$
-~~~$\gtCtx ::= \emptyctx \pipe \gtCtxX{x}{\sigma}$
-\begin{mathpar}
-\inferrule[abs-empty]{ }{
-	\eCtxTogCtxX{\emptyctx}{\emptyctx}
-}
-
-\inferrule[abs-ctx]{
-	\eCtxTogCtxX{\etCtx}{\gtCtx}\\
-	\ddbarX{\trepof{\tau}}{\sigma}
-}{
-	\eCtxTogCtxX{\etCtxX{x}{\tau}}{\gtCtxX{x}{\sigma}}
-}
-\end{mathpar}
-$\fbox{\inferrule{}{\checkRCX{\gamma}{\sigma}}}$
-\begin{mathpar}
-\inferrule[abs-i-var]{ }{
-	\checkRC{\gtCtxX{x}{\sigma}}{\Xi}{\fvalCtx}{\evar{x}}{\sigma}
-}
-
-\inferrule[abs-i-lam]{
-	\ddbarX{\sigma_1}{\sigma_1'}\\
-	\checkRC{\gtCtxX{x}{\sigma_1'}}{\Xi}{\fvalCtx}{\gamma}{\sigma_2}
-}{
-	\checkRCX{\ilam{x}{\sigma_1}{\gamma}}{\darrow{\sigma_1'}{\sigma_2}}
-}
-
-\inferrule[abs-i-ap]{
-	\checkRCX{\gamma_1}{\darrow{\sigma_1}{\sigma_2}}\\
-	\checkRCX{\gamma_2}{\sigma_1}
-}{
-	\checkRCX{\iapp{\gamma_1}{\gamma_2}}{\sigma_2}
-}
-
-\inferrule[abs-i-fix]{
-	\ddbarX{\sigma}{\sigma'}\\
-	\checkRC{\gtCtxX{x}{\sigma'}}{\Xi}{\fvalCtx}{\gamma}{\sigma'}
-}{
-	\checkRCX{\ifix{x}{\sigma}{\gabs}}{\sigma'}
-}
-\\
-\text{\color{gray} (standard statics for integers and products omitted)}
-\\
-%\inferrule[abs int]{ }{
-%	\checkRCX{\iintlit}{\dint}
-%}
-%
-%\inferrule[abs op]{
-%	\checkRCX{\gamma_1}{\dint}\\
-%	\checkRCX{\gamma_2}{\dint}
-%}{
-%	\checkRCX{\iop{\gamma_1}{\gamma_2}}{\dint}
-%}
-%
-%\inferrule[abs pair]{
-%	\checkRCX{\gamma_1}{\sigma_1}\\
-%	\checkRCX{\gamma_2}{\sigma_2}
-%}{
-%	\checkRCX{\ipair{\gamma_1}{\gamma_2}}{\dpair{\sigma_1}{\sigma_2}}
-%}
-%
-%\inferrule[abs fst]{
-%	\checkRCX{\gamma}{\dpair{\sigma_1}{\sigma_2}}
-%}{
-%	\checkRCX{\ifst{\gamma}}{\sigma_1}
-%}
-%
-%\inferrule[abs snd]{
-%	\checkRCX{\gamma}{\dpair{\sigma_1}{\sigma_2}}
-%}{
-%	\checkRCX{\isnd{\gamma}}{\sigma_2}
-%}
-%
-% \inferrule[abs-int-eq]{
-% 	\checkRCX{\gamma_1}{\dint}\\
-% 	\checkRCX{\gamma_2}{\dint}\\\\
-% 	\checkRCX{\gamma_3}{\sigma}\\
-% 	\checkRCX{\gamma_4}{\sigma}
-% }{
-% 	\checkRCX{\iIfEq{\gamma_1}{\gamma_2}{\dint}{\gamma_3}{\gamma_4}}{\sigma}
-% }
-\inferrule[abs-iterm-inverse]{
-	\checkRCX{\gamma}{\sigma}
-}{
-	\checkRCX{\iup{\titerm{\gamma}}}{\sigma}
-}
-
-\inferrule[show-trans]{
-	\fvar{Fam} \in \Xi\\
-	\concrep{\fvalCtx}{\ttype{Fam}{\tauidx}}{\sigma}\\
-	\checkRCX{\gamma}{\sigma}
-}{
-	\checkRCX{\tvalof{\titerm{\gamma}}{\ttype{Fam}{\tauidx}}}{\sigma}
-}
-
-\inferrule[hide-trans]{
-	\fvar{Fam} \notin \Xi\\
-	\concrep{\fvalCtx}{\ttype{Fam}{\tauidx}}{\sigma}\\
-	\checkRCX{\gamma}{\sigma}
-}{
-	\checkRCX{\tvalof{\titerm{\gamma}}{\ttype{Fam}{\tauidx}}}{\trepof{\ttype{Fam}{\tauidx}}}
-}
-\end{mathpar}
-\caption{\small Abstracted internal typing}
-\label{ait}
-\end{figure}
-%
-%family RECORD of (string, type) dict {
-%  new(idx, args. foldl (fn (x, y) => 
-%  get[field : string](idx. case find(idx, field) of SOME t => t | _ => err)
-
-\section{Safety of @$\lambda$}\label{safety}
-In giving users of a language direct influence on the typechecking and translation of expressions, it is essential to consider safety properties.
+%%family RECORD of (string, type) dict {
+%%  new(idx, args. foldl (fn (x, y) => 
+%%  get[field : string](idx. case find(idx, field) of SOME t => t | _ => err)
+%
+%\section{Safety of @$\lambda$}\label{safety}
+%In giving users of a language direct influence on the typechecking and translation of expressions, it is essential to consider safety properties.
 It must not be possible for well-typed programs to fail to finish compiling or go wrong at run-time because of a buggy extension.
@@
 By carefully designing our extension mechanism, we believe we have achieved each of these goals.
-
-\subsection{Representational Consistency and Type Safety}
-We do not need to give a semantics to internal terms and internal types that do not survive deabstraction.
+%
+%\subsection{Representational Consistency and Type Safety}
+%We do not need to give a semantics to internal terms and internal types that do not survive deabstraction.
 The remaining terms and types form a variant of PCF with well-understood primitives (here, integers and binary products) known to be type safe \cite{pfpl}.
@@
 The proof is by straightforward (though deeply nested!) rule induction.
-
-%\begin{theorem}[Representational Consistency]
-%If $\progOK{ }{\Sigma}{e}$ and $\ecompiles{ }{\fvalCtx}{e}{\tau}{\gabs}$ and $\eraseX{\gabs}{\gamma}$ and 	$\ddbar{\Xi_0}{\fvalCtx}{\trepof{\tau}}{\sabs}$ and $\eraseX{\sabs}{\sigma}$ then $\checkRC{ }{\Xi_0}{\fvalCtx}{\gamma}{\sigma}$.
-%\end{theorem}
-
-\subsection{Decidability}
-To show that compilation is decidable, we need to show that both kind checking and active typechecking and compilation are decidable.
+%
+%%\begin{theorem}[Representational Consistency]
+%%If $\progOK{ }{\Sigma}{e}$ and $\ecompiles{ }{\fvalCtx}{e}{\tau}{\gabs}$ and $\eraseX{\gabs}{\gamma}$ and 	$\ddbar{\Xi_0}{\fvalCtx}{\trepof{\tau}}{\sabs}$ and $\eraseX{\sabs}{\sigma}$ then $\checkRC{ }{\Xi_0}{\fvalCtx}{\gamma}{\sigma}$.
+%%\end{theorem}
+%
+%\subsection{Decidability}
+%To show that compilation is decidable, we need to show that both kind checking and active typechecking and compilation are decidable.
 This hinges on showing that the type-level language is typesafe and that evaluation of type-level terms always terminates.
-
-%\begin{theorem}[Kind Safety and Termination]
-%If $\vdash_{\Sigma} \tau : \kappa$ then $\tEvalX{\tau}{\tau'}$ such that $\vdash_{\Sigma_0} \tau' : \kappa$ and $\tEvalX{\tau'}{\tau'}$. 
-%\end{theorem}
-
-The proof is straightforward because the type-level language is based on a conventional simply-typed lambda calculus with only bounded recursion over lists, so standard techniques (e.g.
+%
+%%\begin{theorem}[Kind Safety and Termination]
+%%If $\vdash_{\Sigma} \tau : \kappa$ then $\tEvalX{\tau}{\tau'}$ such that $\vdash_{\Sigma_0} \tau' : \kappa$ and $\tEvalX{\tau'}{\tau'}$. 
+%%\end{theorem}
+%
+%The proof is straightforward because the type-level language is based on a conventional simply-typed lambda calculus with only bounded recursion over lists, so standard techniques (e.g.
 logical relations) can be used directly.
@@
 If the type-level language included constructs like inductive datatypes (without a strict positivity condition) or general recursion, this theorem would be weakened.
-
-\subsection{Non-Interference}
-Our abstraction mechanism guarantees that the representation invariants collectively maintained by the operators associated with a type family cannot be violated by operators in other type families, by ensuring that introductory forms cannot be defined outside the family.
+%
+%\subsection{Non-Interference}
+%Our abstraction mechanism guarantees that the representation invariants collectively maintained by the operators associated with a type family cannot be violated by operators in other type families, by ensuring that introductory forms cannot be defined outside the family.
 One way to state this is as follows:
-
-%\begin{theorem}[Non-Interference]
-%For any property $P(\gamma)$, if $\progOK{ }{\Sigma}{e}$ and $\ecompiles{ }{\fvalCtx}{e}{\tau}{\gabs}$ and $\eraseX{\gabs}{\gamma}$ implies $P(\gamma)$, then for any $\Sigma'$ and $\fvalCtx'$ disjoint from $\Sigma$ and $\fvalCtx$, we have that if $\progOK{ }{\Sigma,\Sigma'}{e'}$ and $\ecompiles{ }{\fvalCtx,\fvalCtx'}{e'}{\tau}{\gabs'}$ and $\eraseX{\gabs'}{\gamma'}$ then $P(\gamma')$.
-%\end{theorem}
-The proof relies on the fact that the show/hide rules are mutually exclusive and so there are two cases to consider for each form of expression, and weakening properties of the family contexts.
+%
+%%\begin{theorem}[Non-Interference]
+%%For any property $P(\gamma)$, if $\progOK{ }{\Sigma}{e}$ and $\ecompiles{ }{\fvalCtx}{e}{\tau}{\gabs}$ and $\eraseX{\gabs}{\gamma}$ implies $P(\gamma)$, then for any $\Sigma'$ and $\fvalCtx'$ disjoint from $\Sigma$ and $\fvalCtx$, we have that if $\progOK{ }{\Sigma,\Sigma'}{e'}$ and $\ecompiles{ }{\fvalCtx,\fvalCtx'}{e'}{\tau}{\gabs'}$ and $\eraseX{\gabs'}{\gamma'}$ then $P(\gamma')$.
+%%\end{theorem}
+%The proof relies on the fact that the show/hide rules are mutually exclusive and so there are two cases to consider for each form of expression, and weakening properties of the family contexts.
 This powerful theorem allows us to compose type families arbitrarily without needing to handle cases in our implementation that are ruled out based on a local analysis.
 The details will be provided in the appendix.
-
-% \subsection{Modular Verification}
-% nope
-
-% that's all folks
-
-\section{Use Cases}\label{examples}
-
-Aldrich argues that dynamic dispatch has proven useful for building  large software frameworks \cite{aldrich-onward13}.
+%
+%% \subsection{Modular Verification}
+%% nope
+%
+%% that's all folks
+%
+%\section{Use Cases}\label{examples}
+%
+%Aldrich argues that dynamic dispatch has proven useful for building  large software frameworks \cite{aldrich-onward13}.
 Object-oriented languages support dynamic dispatch directly but encodings of dynamic dispatch in terms of standard functional primitives can be awkward and difficult to reason about (e.g.
@@
 X).
-
-Indexed type families and indexed operator families are a powerful tool for programming language semantics.
+%
+%Indexed type families and indexed operator families are a powerful tool for programming language semantics.
 In \emph{Practical Foundations for Programming Languages}, for example, nearly every chapter defines a new collection of indexed types and operators and gives their semantics in isolation from the constructs in the other chapters \cite{pfpl}.
@@
 In the appendix, we show how to implement sums and products in \atlam.
-
-The book also shows constructs as varied as dynamic types as well as a number of constructs for parallelism and concurrency, such as futures, promises and actors.
+%
+%The book also shows constructs as varied as dynamic types as well as a number of constructs for parallelism and concurrency, such as futures, promises and actors.
 With a sufficiently capable internal language (exposing basic concurrency primitives, for example), each of these could be fully implemented as libraries using our mechanism.
@@
 In a preliminary implementation of this calculus, Ace, which is beyond the scope of this paper, we have implemented the entirety of the OpenCL programming language as a library, along with primitives supporting partitioned global address spaces.
-
-Object systems too could be implemented using this mechanism, with indices serving to capture the inheritance data and the signatures.
+%
+%Object systems too could be implemented using this mechanism, with indices serving to capture the inheritance data and the signatures.
 Operator families for reading and writing fields and/or sending messages would be parameterized by strings naming them.
@@
 Another related example would be to implement a safe and efficient interoperability layer with an existing OO language by capturing its type system as an extension, to be used only at language boundaries.
-
-Finally, a variety of domain-specific type systems, capturing complex rule systems like the system of scientific units of measure (a family indexed by the measure being used and the type which the measure is being applied to), regular expressions \cite{regexp} (a family indexed by the number of captured groups) or XML \cite{xml} (indexed by a document schema) would be possible.
+%
+%Finally, a variety of domain-specific type systems, capturing complex rule systems like the system of scientific units of measure (a family indexed by the measure being used and the type which the measure is being applied to), regular expressions \cite{regexp} (a family indexed by the number of captured groups) or XML \cite{xml} (indexed by a document schema) would be possible.
 Recent work examined a specialized type system capturing the semantics of the widely-used jQuery javascript library \cite{jquery-oopsla}.
 Today, these all require \emph{ad hoc} language-external solutions, or the development of new languages.
-
-%\section{Design Considerations}\label{design}
-
-\section{Related Work}\label{related-work}
-Our representation schema abstraction mechanism relates closely to abstract and existential types \cite{pfpl,atpl}.
+%
+%%\section{Design Considerations}\label{design}
+%
+%\section{Related Work}\label{related-work}
+%Our representation schema abstraction mechanism relates closely to abstract and existential types \cite{pfpl,atpl}.
 Our calculus enforces the abstraction barriers  in a purely syntactic manner, as in previous work on syntactic type abstraction \cite{syntypeabs}.
@@
 a module), we focus on abstracting away the knowledge of how a primitive type family is implemented outside of a limited scope.
-
-The representational consistency mechanism brings into the language work on typed compilation, especially work done on Standard ML in the TILT project \cite{tilt}.
+%
+%The representational consistency mechanism brings into the language work on typed compilation, especially work done on Standard ML in the TILT project \cite{tilt}.
 Indeed, the specification of Standard ML is structured around a typed internal language and a judgement that assigns a type and an internal term to each expression \cite{smlstd}.
 Representational consistency is related to the notion of type-directed copmilation in this work.
-
-Type-level computation is supported in some form by a growing number of languages.
+%
+%Type-level computation is supported in some form by a growing number of languages.
 For example, Haskell supports a simple form of it \cite{Chakravarty:2005:ATC}).
@@
 We show how to integrate language extensions into the type-level language, drawing on ideas about \cite{activelibraries}.
-
-% \subsection{Run-Time Indirection}
-% {\it Operator overloading} \cite{vanWijngaarden:Mailloux:Peck:Koster:Sintzoff:Lindsey:Meertens:Fisker:acta:1975} and {\it metaobject dispatch} \cite{Kiczales91} are run-time protocols that translate operator invocations into function calls.
+%
+%% \subsection{Run-Time Indirection}
+%% {\it Operator overloading} \cite{vanWijngaarden:Mailloux:Peck:Koster:Sintzoff:Lindsey:Meertens:Fisker:acta:1975} and {\it metaobject dispatch} \cite{Kiczales91} are run-time protocols that translate operator invocations into function calls.
 The function is typically selected according to the type or value of one or more operands.
@@
 However, type-level specification is a {\it compile-time} protocol focused on enabling specialized verification and implementation strategies, rather than simply enabling run-time indirection.
-
-Many languages and tools allow developers to rewrite expressions according to custom rules.
+%
+%Many languages and tools allow developers to rewrite expressions according to custom rules.
 These can broadly be classified as {\it term rewriting systems}.
@@
 MetaML \cite{Sheard:1999:UMS}), and external rewrite systems also exist for many languages.
-These facilities differ from AT\&T in that they involve direct manipulation of terms, while AT\&T involves extending typechecking and translation logic directly.
+%These facilities differ from AT\&T in that they involve direct manipulation of terms, while AT\&T involves extending typechecking and translation logic directly.
 We also draw a distinction between the type-level, used to specify types and compile-time logic, the expression grammar, used to describe run-time behavior, and the internal language, used to implement this behavior.
 By doing so, each component can be structured and constrained as appropriate for its distinct role, as we have shown.
-
-Previous work on extensible languages has suffered from problems with either expressiveness or safety.
+%
+%Previous work on extensible languages has suffered from problems with either expressiveness or safety.
 For example, a number of projects, such as SugarJ \cite{erdweg2011sugarj}, allow for user-defined desugarings (and indeed, our system would clearly benefit from integration with such a mechanism), but this does not allow the semantics to be fundamentally extended nor for implementation details to be hidden.
@@
 Our work aims to provide a sound and safe underpinning, based around well-understood concepts of type and operator families, to language extensibility.
-
-In the future, we will investigate mechanisms to enable type systems with specialized binding and scoping rules, as well as integration of dependent kinds to support mechanized verification of key properties like representational consistency and adequacy against a declarative specification at kind-checking time.
+%
+%In the future, we will investigate mechanisms to enable type systems with specialized binding and scoping rules, as well as integration of dependent kinds to support mechanized verification of key properties like representational consistency and adequacy against a declarative specification at kind-checking time.
 We will also investigate mechanisms that enable a more natural syntax (e.g.
 Wyvern's type-directed syntax \cite{globaldsl}).
-
-%\item With type-level specification, dispatch to a type-level function occurs implicitly on the basis of the structure of an expression.
+%
+%%\item With type-level specification, dispatch to a type-level function occurs implicitly on the basis of the structure of an expression.
 In contrast, most term-rewriting systems operate by  explicit invocation of a macro or specialized syntax.
@@
 A*\cite{fowler2010domain}, EPP\cite{fowler2010domain}) and macro systems for object-oriented languages, like OpenC++ \cite{fowler2010domain} and OpenJava \cite{fowler2010domain}, do offer a somewhat limited form of operation-based dispatch.
-
-% \subsection{Language Frameworks}
-% When the mechanisms available in an existing language prove insufficient, researchers and domain experts must design a new language.
+%
+%% \subsection{Language Frameworks}
+%% When the mechanisms available in an existing language prove insufficient, researchers and domain experts must design a new language.
 A number of tools have been developed to assist with this task, including compiler generators, language workbenches and domain-specific language frameworks (cf \cite{fowler2010domain}).
-
-% A major barrier to adoption is the fact that interoperability is intrinsically problematic.
+%
+%% A major barrier to adoption is the fact that interoperability is intrinsically problematic.
 Even languages which target a common platform, such as the Java Virtual Machine, can only interact using its limited set of primitives.
@@
 Java).
-
-% Instead of focusing on defining standalone languages, type-level specification gives greater responsibility in a granular manner to libraries.
+%
+%% Instead of focusing on defining standalone languages, type-level specification gives greater responsibility in a granular manner to libraries.
 In this way, a range of constructs can coexist within the same program and, assuming that it can be shown by some method that various constructs are safely composable, be mixed and matched.
@@
 Nevertheless, as Ace shows, a simple grammar can be used quite flexibly.
-% \subsection{Extensible Compilers}
-% An alternative methodology is to implement language features granularly as compiler extensions.
+%% \subsection{Extensible Compilers}
+%% An alternative methodology is to implement language features granularly as compiler extensions.
 As discussed in Section 1, existing designs suffer from the same problems related to composability, modularity\-, safety and security as extensible languages, while also adding the issue of language fragmentation.
-
-% Type-level specification can in fact be implemented within a compiler, rather than provided as a core language feature.
+%
+%% Type-level specification can in fact be implemented within a compiler, rather than provided as a core language feature.
 This would resolve some of the issues, as described in this paper.
@@
 We also avoid the fragmentation issue.
-% \subsection{Specification Languages}
-% Several {\it specification languages} (or {\it logical frameworks}) based on these theoretical formulations exist, including the OBJ family of languages (e.g.
+%% \subsection{Specification Languages}
+%% Several {\it specification languages} (or {\it logical frameworks}) based on these theoretical formulations exist, including the OBJ family of languages (e.g.
 CafeOBJ \cite{Diaconescu-Futatsugi01}).
@@
 The  language itself specifies which verification and execution strategies are used.
-
-% Type-determined compilation takes a more concrete approach to the problem, focusing on combining {\it implementations} of different\- logics, rather than simply their specifications.
+%
+%% Type-determined compilation takes a more concrete approach to the problem, focusing on combining {\it implementations} of different\- logics, rather than simply their specifications.
 In other words, it focuses on combining {\it type checkers} and {\it implementation strategies} rather than more abstract representations of a language's type system and dynamic semantics.

### 2014-02-28 att-icfp14/att-icfp14.tex
More icfp progress

@@
 \section{Introduction}
-Typed programming languages are often described in fragments, each consisting of a small number of type constructors (often one) and associated (term-level) operator constructors.
+Typed programming languages are often described in fragments, each consisting of a small number of indexed type constructors (often one) and associated (term-level) operator constructors.
 The simply typed lambda calculus (STLC), for example, consists of a single fragment containing a single type constructor, $\rightarrow$, indexed by a pair of types, and two operator constructors: $\lambda$, indexed by a type, and \verb|ap|, which we may think of as being indexed trivially.
@@
 parametric polymorphism), we might take pause.
-In $\mathcal{L}\{\rightarrow\,\forall\}$, studied variously by Girard as System $\mathbf{F}$ \cite{girardF} and Reynolds as the polymorphic lambda calculus \cite{polylam}, it is known that sums, products, and inductive and co-inductive types can all be weakly defined via a technique similar to Church encodings \cite{reynolds}\todo{what to cite for this?}.
-While of substantial theoretical (and pedagogical) interest, this fact does not render irrelevant the construction of a language like $\mathcal{L}\{\rightarrow\forall~\mathtt{nat} + \times\}$.
-Reynolds, in a remark that recalls the ``Turing tarpit'' of Perlis \cite{Perl82a}, cautioned against using the existence of a weak encoding as an argument for the practicality of programming with the corresponding embedding directly \cite{Reynolds94anintroduction}:
+In $\mathcal{L}\{\rightarrow\,\forall\}$, studied variously by Girard as System $\mathbf{F}$ \cite{girardF} and Reynolds as the polymorphic lambda calculus \cite{polylam}, it is known that sums, products, and inductive and co-inductive types can all be weakly defined via Church-style encodings \cite{reynolds}\todo{what to cite for this?}.
+This fact is of substantial theoretical importance and is quite relevant if \emph{translating} a language like $\mathcal{L}\{\rightarrow\forall~\mathtt{nat} + \times\}$ to $\mathcal{L}\{\rightarrow\forall\}$ is one's main interest.
+But Reynolds, in a remark that recalls the ``Turing tarpit'' of Perlis \cite{Perl82a}, cautioned against using the existence of a weak encoding as an argument for programming with the corresponding embedding directly \cite{Reynolds94anintroduction}:
 \begin{quote}
@@
 
-Finding a weak encoding of a language fragment in terms of a collection of others is useful if \emph{implementing the dynamic semantics correctly} is one's main interest, but adding new fragments directly to a language can make statically reasoning about programs more precise, support a more natural programming style and endow the language with a more favorable cost semantics.
-Consistent with this view, typed programming languages like ML expose, for example, both $n$-ary product types and record types, despite the fact that any particular product or record type is weakly definable using  polymorphic functions (or simply in terms of the other).
-The compiler for such a language, on the other hand, \emph{is} mainly concerned with implementing the dynamic semantics correctly, and indeed, compilers often use the fact that weak definability results are quite readily derivable to their advantage, translating the constructs of their external language (EL) to those of a much simpler internal language (IL) after typechecking is complete.
+%Adding new fragments directly to a language can make statically reasoning about programs more precise, support a more natural programming style and endow the language with a more favorable cost semantics.
+The latter two points might be relevant even if a strong embedding (i.e.
+one where there is a semantics-preserving isomorphism between terms and types of the fragment and the language) can be found.
+%
+%Consistent with this view, typed programming languages like ML expose, for example, both $n$-ary product types and record types, despite the fact that any product type is definable using a record type (or in the other direction, using a module to provide the field selection operator).
+The compiler, on the other hand, \emph{is} mainly concerned with implementing the dynamic semantics correctly, and indeed, compilers often use the fact that weak definability results are quite readily derivable to their advantage, translating the constructs of an external language (EL) to those of a much simpler internal language (IL), e.g.
+with only binary products, during or directly after typechecking\todo{cite Harper-Stone}.
 %Correctly implementing the dynamic semantics of the language becomes the primary concern during this phase, so a simple IL that decreases the number of cases needing consideration when implementing and verifying the compiler is a wise choice.
 
-Having established why a minimal language like $\mathcal{L}\{\rightarrow\forall\}$, while suitable as an internal language, needs to be extended with additional fragments before it is suitable for use as a human-facing external language, we might now ask whether modern ``general-purpose'' languages like ML are fundamentally different.
-These languages have been used successfully for a wide variety of computing tasks, admitting many more strong encodings than a  calculus like $\mathcal{L}\{\rightarrow \forall\}$.
-It is certainly not our intention to argue against the notion that the particular fragments they have chosen to privilege occupy ``sweet spots'' in the design space.
-But in fact, situations continue to arise in both research and practice where the direct introduction of a new fragment is desirable because it can still only be weakly defined in terms of existing fragments, or a strong encoding is possible but impractically verbose or inefficient:
+%Having established why a minimal language like $\mathcal{L}\{\rightarrow\forall\}$, while suitable as an internal language, needs to be extended with additional fragments before it is suitable for use as a human-facing external language,
+``General-purpose'' languages like ML, on the other hand, expose mechanisms that admit strong embeddings of many more fragments and occupy what is widely seen as a ``sweet spot'' in the design space.
+Nevertheless, situations continue to arise in both research and practice where introducing a new fragment would be desirable because an abstraction can still only be weakly defined in terms of existing fragments, or a strong embedding, while possible, is impractically verbose or inefficient:
 \begin{enumerate}
 \item General-purpose abstractions continue to evolve and admit variants.
-There are many  variations on record types, for example: records with functional record update, prototypic delegation or specified field ordering (that is, labeled tuples) are either awkward to strongly define or cannot be.
+There are many  variations on record types, for example: records with functional record update, specified field ordering (that is, labeled tuples) or prototypic delegation are either awkward to strongly embed or cannot be.
 Similarly, sum types also admit variants (e.g.
 various forms of open sums).
-Even something as seemingly simple as a type-safe \verb|printf| operator requires extending a language like Ocaml  \cite{ocaml-printf}.
+Even something as seemingly simple as a type-safe \verb|sprintf| operator requires extending a language like Ocaml  \cite{ocaml-printf}.
 
 \item Perhaps more interestingly, specialized type systems that enforce stronger invariants than general-purpose constructs are capable of enforcing are often developed by researchers.
-One need take only a brief  excursion through the literature to discover language extensions that support parallel programming \cite{a}\todo{citations}, concurrent programming \cite{cml}, distributed programming \cite{tom7}, dataflow programming \cite{reactiveml}, authenticated data structures \cite{popl13}, information flow security \cite{walker00, smith2001}, database queries \cite{db}, aliased references \cite{naden12}, network protocols \cite{sekar99}, units of measure \cite{keneddy} and a bounty of others.% All of these are implemented as dialects of existing languages, presumably because a strong encoding was not feasible.
+One need take only a brief  excursion through the literature to discover language extensions that support parallel programming \cite{a}\todo{citations}, concurrent programming \cite{cml}, distributed programming \cite{tom7}, dataflow programming \cite{reactiveml}, authenticated data structures \cite{popl13}, information flow security \cite{walker00, smith2001}, database queries \cite{db}, aliased references \cite{naden12}, network protocols \cite{sekar99}, units of measure \cite{keneddy} and many others.% All of these are implemented as dialects of existing languages, presumably because a strong encoding was not feasible.
 \item Foreign function interfaces (FFIs) that allow programmers to safely and naturally interact with libraries written in an existing language require enforcing the type system of the foreign language within the calling language.
@@
 
-These language fragments are generally disseminated as distinct \emph{dialects} of an existing general-purpose language, constructed either as a fork, a feature behind a compiler flag or, using tools like compiler generators, DSL frame\-works or  language workbenches.
+These sorts of innovations are generally disseminated as distinct \emph{dialects} of an existing general-purpose language, constructed either as a fork, using tools like compiler generators, DSL frame\-works or  language workbenches, or directly within a particular compiler for the language, sometimes activated by a flag or pragma.
 This, we argue, is quite unsatisfying: a programmer can choose either a dialect supporting an innovative approach to parallel programming or one that builds in support for statically reasoning about units of measure, but there may not be an available dialect supporting both.
 Forming such a dialect is alarmingly non-trivial, even in the rare situation where a common framework has been used or the dialects are implemented within the same compiler, as these mechanisms do not guarantee that different combinations of individually sound language fragments remain sound when combined.
-Metatheoretic and compiler correctness results can only be derived for the dialect \emph{resulting} from language   composition operations, so even  fragmentary frameworks induce an unbounded, combinatoric collection of proof obligations that providers have little choice but to  burden clients with.% We will consider this further in our discussion of related work in Sec.
+Metatheoretic and compiler correctness results can only be derived for the dialect \emph{resulting} from a language   composition operation, so fragment providers have little choice but to leave this to clients.% We will consider this further in our discussion of related work in Sec.
 \ref{related-work}.
@@
 Well-designed languages preclude the  possibility of conflict between two libraries and ensure that the semantics of one library  cannot be weakened by another by strictly enforcing abstraction barriers.
-For example, a module declaring an abstract type in ML can rely on its representation invariants no matter which other modules are in use, so clients can assume that they will operate robustly in combination without attending to burdensome ``link-time'' proof obligations.
-
-We draw inspiration from this approach to design a mechanism that makes it possible to extend the static and dynamic semantics of an external portion of the language with new type and operator constructors from within.
-The semantics of the language as a whole constrains these extensions to maintain type safety and decidability of type checking and type equality.
-Moreover, an important class of lemmas that play a key role in  proofs that an extension enables strong encodings of a language fragment can be derived assuming a ``closed world'' (that is, in the completely traditional manner where one constructs a complete language and reasons inductively).
-The language  guarantees that they will be conserved in the  ``open world'' by enforcing  representation independence at extension boundaries.
-
-Our mechanism essentially tasks the type-level language, in a controlled manner, with implementing portions of the logic that would normally be implemented in the front-end of the compiler.
-When typechecking a user-defined operator invocation, the semantics invoke a type-level function associated with the operator's constructor to \emph{compute} both a type and a translation to a fixed typed internal language (together called a \emph{derivate}).
-Language fragments that can be weakly defined in terms of the typed internal language can be made strongly definable in the external language by introducing, in this way, the necessary  constructors and static semantics.
-We call this an \emph{actively typed semantics}.
+For example, a module declaring an abstract type in ML can rely on its representation invariants no matter which other modules are in use, so clients can assume that they will operate robustly in combination without needing to attend to burdensome ``link-time'' proof obligations.
+
+\paragraph{Contributions}
+We draw inspiration from this approach to design a mechanism that makes it possible to declare new type and operator constructors and define,  {using type-level functions}, their static and dynamic semantics, the latter by translation to a fixed typed internal language, from within.
+If a language fragment can be weakly defined in terms of the typed internal language and its statics are of a form that can be implemented by this protocol, it can be made strongly embeddable in the external language by introducing, in this way, the necessary constructors and logic.
+
+The protocol constrains extensions to maintain type safety and decidability of type checking and type equality.
+Moreover, an important class of lemmas that play a key role in  proofs that an extension enables strong embeddings of a language fragment can be derived assuming a ``closed world'' (that is, in the completely traditional manner where one constructs a complete language and reasons inductively).
+The semantics guarantees that they will be conserved in the  ``open world'' by enforcing a form of  representation independence at all extension boundaries.
+Thus, if a strong embedding can be constructed using some combination of extensions, it can be constructed in the presence of arbitrary combinations of other extensions as well, avoiding the problems faced by language dialects and justifying the inclusion of the mechanism directly into the language.
 
 We will begin in Sec.
-\ref{atlam} by introducing a minimal calculus with an actively typed semantics, @$\lambda$.
-The external language builds in only the $\rightarrow$ fragment.
-We then use the $\mathtt{nat}$ fragment of G\"odel's $\mathbf{T}$ and a fragment defining $n$-ary products as pedagogical examples to explain how extensions are written and motivate the constraints that the semantics must impose to make the guarantees that we claim.
-We then examine the semantics in greater detail and state and sketch proofs of strong safety and conservativity theorems in Sec.
+\ref{core} by introducing a minimal calculus with an actively typed semantics, @$\lambda$.
+The external language builds in only the $\rightarrow$ fragment and the typed internal language is a simple variant of PCF that permits only weak embeddings of even simple fragments like the $\mathtt{nat}$ fragment of G\"odel's $\mathbf{T}$ or ML-style $n$-ary products.
+We implement these as extensions to explain the mechanism and motivate the constraints that the semantics imposes to make the guarantees that we claim.
+We then examine the semantics in greater detail and state and describe the key points in the proofs of the metatheory in Sec.
 \ref{theory}.
-\todo{organization of rest of sections}
-
-\section{@$\lambda$}
-\begin{figure*}[t]
-\small
-$$\begin{array}{rccl}	
-\textbf{programs} & \rho & ::= & \pfam{\familyDf}{\progsort} 
-%\pipe \pdef{t}{\kappa}{\tau}{\progsort} 
-\pipe e\\
-		&	\theta	&	::= &	\tops{op}{\kappaidx}{i}{a}{\taut{def}} \pipe 
-												\topp{\theta}{\theta}\\
-\\
-\textbf{external terms} 				&	e	&	::=	&	\evar{x} \pipe 
-%														\efix{x}{\tau}{e} \pipe 
-														\elam{\evar{x}}{\tau}{e} \pipe 
-														\eop{Tycon}{op}{
-															\tauidx
-														}{
-  												    		\splat{e}{1}{n}
-														} \\
-									& 		&		& 	\\
-
-\textbf{internal terms} 				& 	\iota	&	::=	&	\evar{x} \pipe 
-												\ifix{\evar{x}}{\sigma}{\iota} \pipe
-												\ilam{\evar{x}}{\sigma}{\iota} \pipe 
-												\iapp{\iota_{1}}{\iota_{2}} \pipe
-												\iintlit \pipe \iop{\iota_{1}}{\iota_{2}} \pipe \iIfEq{\iota_{1}}{\iota_{2}}{\dint}{\iota_{3}}{\iota_{4}} \\
-										& & \pipe & 
-												\iunit \pipe
-												\ipair{\iota_{1}}{\iota_{2}} \pipe 
-												\ifst{\iota} \pipe
-												\isnd{\iota} \pipe
-												 \iinl{\sigma_2}{\iota_1} \pipe \iinr{\sigma_1}{\iota_2} \pipe \icase{e}{x}{e_1}{x}{e_2} \\
-
-%\text{deabstracted}& \iota & ::= & \mathcal{G}[\iota, \sigma]\\
-\textbf{internal types}			&	\sigma	&	::=	&    \darrow{\sigma_1}{\sigma_2} \pipe \dint \pipe \dunit \pipe \dpair{\sigma_1}{\sigma_2} \pipe \dsum{\sigma_1}{\sigma_2}
-
-\\
-\\
-
-\hspace{-5pt}\textbf{type-level terms} 	& \tau 	& ::= 	& 	\tvar{t} \pipe 
-														\tlam{t}{\kappa}{\tau} \pipe 
-														\tapp{\tau_1}{\tau_2} \pipe
-														\tnil{\kappa} \pipe \tcons{\tau_1}{\tau_2} \pipe 
-									                     \tfold{\tau_1}{\tau_2}{h}{t}{r}{\tau_3}
-														\\
-
-	 			& 		& \pipe	& 	 \iintlit \pipe \iop{\tau_1}{\tau_2} \pipe \tlabel{label} \pipe \tunit \pipe 
-														\tpair{\tau_{1}}{\tau_{2}} \pipe 
-														\tfst{\tau} \pipe 
-														\tsnd{\tau} 
-														\\	
-									&       & \pipe & \tinl{\kappa_2}{\tau_1} \pipe \tinr{\kappa_1}{\tau_2} \pipe \tsumcase{\tau}{x}{\tau_1}{x}{\tau_2} \\
-\text{equality}  & & \pipe & 					\tifeq{\tau_{1}}{\tau_{2}}{\kappa}{\tau_{3}}{\tau_{4}} 
-														\\													\text{types} 						& 		& \pipe	& 	\ttypestd \pipe \tfamcase{\tau}{Tycon}{x}{\tau_1}{\tau_2}\\
-						%				& & \pipe & \tfamcase{\tau}{Fam}{x}{\tau_1}{\tau_2}\\
-
-\text{derivates} 				& 		 & 	\pipe	&	\tden{\tauiterm}{\tautype}  \pipe \tderbind{x}{t}{\tau}{\tau_1}\\% \tdencase{\tau}{x}{t}{\tau_1}{\tau_2}\\
- %& & \pipe & 
-%														\tdencase{\tau}{y}{x}{\tau_1}{\tau_2}
-%														 \\
-
-\text{reified IL}		&		&	\pipe	&	\titerm{\bar \iota} \pipe \titype{\bar \sigma} \\
-% &  & \pipe & \tvalof{\tau_1}{\tau_2} \pipe \iup{\tau} \\
-% 												\trepof{\tau} \pipe \dup{\tau}\\
-	& \bar{\iota} & ::= & x \pipe \ifix{x}{\bar \sigma}{\bar \iota} 
-	%\pipe \ilam{x}{\bar \sigma}{\bar \iota} \pipe \iapp{\bar \iota_1}{\bar \iota_2} 
-	\pipe \cdots \pipe \iup{\tau} \pipe \tvalof{\tauiterm}{\tautype} \\						
- & \bar{\sigma} & ::= & \darrow{\bar \sigma_1}{\bar \sigma_2} \pipe \cdots \pipe \dup{\tau} \pipe \trepof{\tau} \\
-											\\
-\textbf{kinds} 					& \kappa	&	::=	&	\karrow{\kappa_1}{\kappa_2} \pipe \klist{\kappa} \pipe \kint \pipe
-											    \klabel \pipe
-												\kunit \pipe 
-												\kpair{\kappa_{1}}{\kappa_{2}} \pipe 
-												\ksum{\kappa_1}{\kappa_2} \pipe
-												\kTypeBlur \pipe \kDen \pipe 
-												\kITerm \pipe \kIType
-												\\
-\\												
-%\textbf{ops signature}			& \Theta	&	::=	&	\kOpEmpty \pipe \kOp{\Theta}{op}{\kappai}\\
-%											 							&		&		&	\\
-\end{array}$$
-\vspace{-10pt}
-\caption{\small Syntax of Core \atlam.
-Here, $x$ ranges over external and internal language variables, $\tvar{t}$ ranges over type-level variables, $\fvar{Tycon}$ ranges over type constructor names, $\opvar{op}$ ranges over operator constructor names, $\iintlit$ ranges over integer literals, $\tlabel{label}$ ranges over label literals (see text) and $\oplus$ ranges over the standard total binary operations over integers (e.g.
-addition).
-The form $\tvalof{\tauiterm}{\tautype}$ is  used internally by the semantics (it need not be supported by a parser).%The productions related to the internal language are written using generators $\mathcal{G}$ and $\mathcal{S}$ to avoid duplicating the syntax of common terms.
-\vspace{-10pt}
-\label{grammar}}
-\end{figure*}
-\begin{figure}[t]
-\small
-\begin{flalign}
-\label{natfam}&\family{Nat}{\kunit}{\\
-\label{z}&\quad\tops{z}{\kunit}{i}{a}{\tlam{idx}{\kunit}{\tlam{args}{\klist{\kDen}}{
-	\tapp{\tapp{\tvar{if\_empty}}{\tvar{args}}}{
-		\tden{\titerm{0}}{\ttype{Nat}{\tunit}}
-	}}}};\\
-\label{s}&\quad\tops{s}{\kunit}{i}{a}{\tlam{idx}{\kunit}{\tlam{args}{\klist{\kDen}}{\\
-& \quad\quad
-	\tapp{
-		\tapp{\tvar{pop\_final}}{\tvar{args}}}{\tlam{x}{\kITerm}{\tlam{t}{\kTypeBlur}{
-			\\
-			&\quad\quad\tapp{\tapp{\tapp{\tvar{check\_type}}{\tvar{t}}}{\ttype{Nat}{\tunit}}}{
-				\tden{\titerm{\iup{\tvar{x}}+1}}{\ttype{Nat}{\tunit}}
-			}
-		}}
-		}
-	}}};\\
-\label{rec}&\quad\tops{rec}{\kunit}{i}{args}{\tlam{idx}{\kunit}{\tlam{args}{\klist{\kDen}}{\\
-& \quad\quad
-	\tapp{\tapp{\tvar{pop}}{\tvar{args}}}{\tlam{x1}{\kITerm}{\tlam{t1}{\kTypeBlur}{\tlam{args}{\klist{\kDen}}{\\
-	&\quad\quad\tapp{\tapp{\tvar{pop}}{\tvar{args}}}{\tlam{x2}{\kITerm}{\tlam{t2}{\kTypeBlur}{\tlam{args}{\klist{\kDen}}{\\
-	&\quad\quad \tapp{\tapp{\tvar{pop\_final}}{\tvar{args}}}{\tlam{x3}{\kITerm}{\tlam{t3}{\kTypeBlur}{\\
-	&\quad\quad \tapp{\tapp{\tapp{\tvar{check\_type}}{\tvar{t1}}}{\ttype{Nat}{\tunit}}}{(\\
-	&\quad\quad \tapp{\tapp{\tapp{\tvar{check\_type}}{\tvar{t3}}}{\ttype{Arrow}{(\ttype{Nat}{\tunit},\ttype{Arrow}{(\tvar{t2},\tvar{t2})})}}}{\\
-		\label{fix}&\quad\quad \tden{\titerm{\iapp{(\ifix{f}{\darrow{\dint}{\trepof{\tvar{t2}}}}{\ilam{x}{\dint}{\\
-		\label{lastop}&\quad\quad\quad \iIfEq{x}{0}{\dint}{\iup{\tvar{x2}}}{\iapp{\iapp{\iup{\tvar{x3}}}{(x-1)}}{(\iapp{f}{(x-1)})}}}})}{\iup{\tvar{x1}}}}}{\tvar{t2}}
-	)}}
-	}}}}}}}}}}}}}}
-\\
-&}{XXX}{i}{\tlam{idx}{\kunit}{\titype{\dint}}};\\
-\label{nattype}&\etdef{nat}{\kTypeBlur}{\ttype{Nat}{\tunit}}{\\
-& \elet{plus}{\ttype{Arrow}{(\tvar{nat}, \ttype{Arrow}{(\tvar{nat}, \tvar{nat})})}}{
-	\elam{x}{\tvar{nat}}{
-		\elam{y}{\tvar{nat}}{
-			\\&\quad \eop{Nat}{rec}{\tunit}{x; y; \elam{p}{\tvar{nat}}{\elam{r}{\tvar{nat}}{
-				\eop{Nat}{s}{\tunit}{r}
-			}}}
-		}
-	}
-}{\\
-& \elet{two}{\tvar{nat}}{\eop{Nat}{s}{\tunit}{\eop{Nat}{s}{\tunit}{\eop{Nat}{z}{\tunit}{}}}}{\\
-& \eop{Arrow}{ap}{\tunit}{plus; \eop{Arrow}{ap}{\tunit}{two; two}}
-}}}
-%{\\&\elet{two}{\tvar{nat}}{\eop{Nat}{s}{\tunit}{\eop{Nat}{s}{\tunit}{\eop{Nat}{z}{\tunit}{ }}}}{\eapp{\eapp{plus}{two}}{two}}}
-%}
-%\elam{x}{\tvar{nat}}{\elam{y}{\tvar{nat}}{\\
-	%&\quad \eop{Nat}{rec}{\tunit}{x; y; \elam{p}{\tvar{nat}}{\elam{r}{\tvar{nat}}{
-	%\eop{Nat}{s}{\tunit}{r}
-	%}}}}
-\end{flalign}
-\vspace{-15px}
-\caption{\small G\"odel's $\mathbf{T}$ in \atlam, used to calculate 2+2.
-The simple helper functions $\tvar{if\_empty}$, $\tvar{pop}$, $\tvar{pop\_final}$, $\tvar{check\_type}$ all have return kind $\ksum{\kDen}{\kunit}$ (see text).
-Their definitions can be found in the appendix.
-We use \textsf{let} to bind both external and type-level variables for clarity of presentation; these can be eliminated by manually performing the indicated substitutions (or added to the semantics).}
-\label{nat}
-\end{figure}
+These two sections contain the most fundamental contributions of this work.
+
+While the core calculus addresses the issue of making a strong embedding possible, embeddings into the core calculus are quite clearly impractical syntactically.
+To begin to address the issue of syntax, we develop in Sec.
+\ref{expanded-syntax} a flexible type-directed dispatch protocol for a more conventional \emph{expanded syntax} that defers to the core calculus semantically.
+We then use the expanded syntax to discuss several more interesting examples in Sec.
+\ref{examples}: \todo{list these}.
+\todo{organization of rest of sections}.
+
+While of surprising expressiveness, @$\lambda$ itself is certainly not capable of expressing the full variety of type systems listed above.
+Indeed, to guarantee safety and conservativity, the calculus is perhaps too conservative.
+For example, it only admits fragments for which the typing judgement looks essentially like the typing judgement of the $\rightarrow$ fragment; new contexts cannot be introduced.
+We discuss this and some of the other reasons why a fragment may not be expressible in @$\lambda$ in Sec.
+\ref{limitations}.
+These limitations do not appear to be fundamental to the approach, so we also outline in Sec.
+\ref{limitations} promising directions for future work.
+We conclude with a discussion of related work in Sec.
+\ref{related-work}.
+
+\section{Core @$\lambda$}\label{core}
+
 %
@@
 %\end{figure}
-In this section, we will develop an ``actively typed'' version of the simply-typed lambda calculus with simply-kinded type-level computation called @$\lambda$.
-More specifically, the level of types, $\tau$, will itself form a simply-typed lambda calculus.
-\emph{Kinds} classify type-level terms in the same way that types conventionally classify expressions.
-Types become just one kind  of type-level value (which we will write $\kTypeBlur$, though it is also variously written $\star$, \verb|T| and \verb|Type| in various settings).
-Rather than there being a fixed set of type constructors, we allow the programmer to declare new type  constructors, and give the static and dynamic semantics of their associated operators, by writing type-level functions.
-In the semantics for this calculus, our kind system combined with techniques borrowed from the typed compilation literature and a form of type abstraction allow us to prove strong type safety, decidability and conservativity theorems.
-
-The syntax of @$\lambda$ is given in Fig.
+
+The syntax of Core @$\lambda$ is given in Fig.
 \ref{grammar}.
-An example of a program containing an extension that implements the static and dynamic semantics of G\"odel's \textbf{T} is given in Fig.
+An example of a program defining type and operator constructors that can be used to strongly embed G\"odel's \textbf{T} in @$\lambda$ is given in Fig.
 \ref{nat}.
-We do not, of course, deny that natural numbers can be strongly encoded with a similar usage and asymptotic performance profile (but with potentially relevant additional function call overhead) by existing means (e.g.
-by an abstract type).
-We will provide more sophisticated examples where this is not the case below.
+We will discuss its semantics and how precisely the strong embedding seen being used in this example works as we go on.
+Natural numbers can, of course, be strongly embedded in existing languages with a similar usage and asymptotic performance profile (up to function call overhead with an abstract type, for example).
+We will provide more sophisticated examples where this is less feasible later on.
 %It consists of:
@@
 \subsection{Programs}
-
-A program, $\rho$, consists of a series of static declarations followed by an external term, $e$.
-The syntax for external terms contains three core forms: variables, functions, and a form for all other operator invocations, which we will discuss below.
-Compiling a program consists of first \emph{kind checking} its static declarations and type-level terms then \emph{type checking} the external term and, simultaneously, \emph{translating} it to a term, $\iota$, in a \emph{typed internal language}.
-In our calculus, the internal language contains partial functions (via the generic fixpoint operator of Plotkin's PCF \cite{PCF}), simple product and sum types and a base type of integers (to make our example interesting and give a nod to practicality on contemporary machines).
-In practice, the internal language could be any intermediate language for which type safety and decidability of typechecking are known, as @$\lambda$ requires these to establish the corresponding theorems about the language as a whole.
-
-The \emph{active typing judgement} relates an external term to a \emph{type} and an internal term, called its \emph{translation}, under \emph{typing context} $\Gamma$ and \emph{constructor context} $\Phi$:
+\begin{figure}[t]
+\small
+$$\begin{array}{rccl}	
+\textbf{programs} & \rho & ::= & \pfam{\familyDf}{\progsort}
+\\& &  \pipe & 
+%\pipe \pdef{t}{\kappa}{\tau}{\progsort} 
+ e\\
+		&	\theta	&	::= &	\tops{op}{\kappaidx}{i}{a}{\taut{def}} \pipe 
+												\topp{\theta}{\theta}\\
+\\
+\textbf{external terms} 				&	e	&	::=	&	\evar{x} \pipe 
+%														\efix{x}{\tau}{e} \pipe 
+														\elam{\evar{x}}{\tau}{e} \pipe 
+														\eop{Tycon}{op}{
+															\tauidx
+														}{
+  												    		\splat{e}{1}{n}
+														} \\
+									& 		&		& 	\\
+
+\textbf{internal terms} 				& 	\iota	&	::=	&	\evar{x} \pipe 
+												\ifix{\evar{x}}{\sigma}{\iota} \pipe
+												\ilam{\evar{x}}{\sigma}{\iota} \pipe 
+												\iapp{\iota_{1}}{\iota_{2}} 
+\\\text{integers}&&\pipe&												
+												\iintlit \pipe \iop{\iota_{1}}{\iota_{2}} \pipe \iIfEq{\iota_{1}}{\iota_{2}}{\dint}{\iota_{3}}{\iota_{4}} \\
+\text{products}										& & \pipe & 
+												\iunit \pipe
+												\ipair{\iota_{1}}{\iota_{2}} \pipe 
+												\ifst{\iota} \pipe
+												\isnd{\iota} 
+\\\text{sums}&&\pipe&												
+												 \iinl{\sigma_2}{\iota_1} \pipe \iinr{\sigma_1}{\iota_2} \\&&\pipe& \icase{e}{x}{e_1}{x}{e_2} \\
+
+%\text{deabstracted}& \iota & ::= & \mathcal{G}[\iota, \sigma]\\
+\textbf{internal types}			&	\sigma	&	::=	&    \darrow{\sigma_1}{\sigma_2} \pipe \dint \pipe \dunit \pipe \dpair{\sigma_1}{\sigma_2} \pipe \dsum{\sigma_1}{\sigma_2}
+
+\\
+\\
+
+\hspace{-5pt}\textbf{type-level terms} 	& \tau 	& ::= 	& 	\tvar{t} \pipe 
+														\tlam{t}{\kappa}{\tau} \pipe 
+														\tapp{\tau_1}{\tau_2} \pipe \iintlit \pipe \iop{\tau_1}{\tau_2} \pipe \tlabel{label}
+\\\text{lists}&&\pipe&											
+														\tnil{\kappa} \pipe \tcons{\tau_1}{\tau_2} \pipe 
+									                     \tfold{\tau_1}{\tau_2}{h}{t}{r}{\tau_3}
+														\\
+
+\text{products}	 			& 		& \pipe	& 	  \tunit \pipe 
+														\tpair{\tau_{1}}{\tau_{2}} \pipe 
+														\tfst{\tau} \pipe 
+														\tsnd{\tau} 
+														\\	
+\text{sums}									&       & \pipe & \tinl{\kappa_2}{\tau_1} \pipe \tinr{\kappa_1}{\tau_2} \\&&\pipe& \tsumcase{\tau}{t}{\tau_1}{t}{\tau_2} \\
+						\text{types} 						& 		& \pipe	& 	\ttypestd \\&&\pipe& \tfamcase{\tau}{Tycon}{x}{\tau_1}{\tau_2}\\
+\text{equality}  & & \pipe & 					\tifeq{\tau_{1}}{\tau_{2}}{\kappa}{\tau_{3}}{\tau_{4}} 
+														\\													%				& & \pipe & \tfamcase{\tau}{Fam}{x}{\tau_1}{\tau_2}\\
+
+\text{derivates} 				& 		 & 	\pipe	&	\tden{{\bar\iota}}{\tau} \pipe \tden{\ibar}{\tau}^{\checkmark} \pipe \ttypeof{\tau} \\% \tdencase{\tau}{x}{t}{\tau_1}{\tau_2}\\
+ %& & \pipe & 
+%														\tdencase{\tau}{y}{x}{\tau_1}{\tau_2}
+%														 \\
+
+\text{rep types}		&		&	\pipe	&	\titype{\bar \sigma} \\
+% &  & \pipe & \tvalof{\tau_1}{\tau_2} \pipe \iup{\tau} \\
+% 												\trepof{\tau} \pipe \dup{\tau}\\
+\text{abstract IL}	& \bar{\iota} & ::= & x \pipe \ifix{x}{\bar \sigma}{\bar \iota} 
+	%\pipe \ilam{x}{\bar \sigma}{\bar \iota} \pipe \iapp{\bar \iota_1}{\bar \iota_2} 
+	\pipe \cdots \pipe \itransof{\tau} \\						
+ & \bar{\sigma} & ::= & \darrow{\bar \sigma_1}{\bar \sigma_2} \pipe \cdots \pipe \dup{\tau} \pipe \trepof{\tau} \\
+											\\
+\textbf{kinds} 					& \kappa	&	::=	&	\karrow{\kappa_1}{\kappa_2} \pipe \kint \pipe
+											    \klabel \pipe
+											    \klist{\kappa} \pipe
+												\kunit \pipe 
+												\kpair{\kappa_{1}}{\kappa_{2}} \\
+												&&\pipe&
+												\ksum{\kappa_1}{\kappa_2} \pipe
+												\kTypeBlur \pipe \kDen \pipe 
+												\kIType								
+%\textbf{ops signature}			& \Theta	&	::=	&	\kOpEmpty \pipe \kOp{\Theta}{op}{\kappai}\\
+%											 							&		&		&	\\
+\end{array}$$
+%\vspace{-10pt}
+\caption{\small Syntax of Core \atlam.
+Here, $x$ ranges over external and internal language variables, $\tvar{t}$ ranges over type-level variables, $\fvar{Tycon}$ ranges over type constructor names, $\opvar{op}$ ranges over operator constructor names, $\iintlit$ ranges over integer literals, $\tlabel{label}$ ranges over label literals (see text) and $\oplus$ ranges over standard total binary operations over integers (e.g.
+addition).%The productions related to the internal language are written using generators $\mathcal{G}$ and $\mathcal{S}$ to avoid duplicating the syntax of common terms.
+\vspace{-10pt}
+\label{grammar}}
+\end{figure}
+\begin{figure}[t]
+\small
+\begin{flalign}
+\label{natfam}&\family{Nat}{\kunit}{\\
+\label{z}&\quad\tops{z}{\kunit}{i}{a}{\tlam{i}{\kunit}{\tlam{a}{\klist{\kDen}}{
+	\tapp{\tapp{\tvar{arity0}}{\tvar{a}}}{\\&\quad\quad
+		\tden{0}{\ttype{Nat}{\tunit}}
+	}}}};\\
+\label{s}&\quad\tops{s}{\kunit}{i}{a}{\tlam{i}{\kunit}{\tlam{a}{\klist{\kDen}}{\tvar{arity1}~\tvar{a}~\tlam{d}{\kDen}{\\
+& \quad\quad \tvar{ifeq}~\ttypeof{\tvar{d}}~\ttype{Nat}{\tunit}\\
+& \quad\quad\quad \tden{\itransof{\tvar{d}} + 1}{\ttype{Nat}{\tunit}}
+	}}}};\\
+\label{rec}&\quad\tops{rec}{\kunit}{i}{args}{\tlam{i}{\kunit}{\tlam{a}{\klist{\kDen}}{\tvar{arity3}~\tvar{a}~\tlam{d1}{\kDen}{\tlam{d2}{\kDen}{\tlam{d3}{\kDen}{\\
+&\quad\quad \tvar{ifeq}~\ttypeof{\tvar{d1}}~\ttype{Nat}{\tunit}\\
+&\quad\quad {\sf let}~\tvar{t2} = \ttypeof{\tvar{d2}}~{\sf in}\\
+&\quad\quad \tvar{ifeq}~\ttypeof{\tvar{d3}}~\ttype{Arrow}{(\ttype{Nat}{\tunit}, \ttype{Arrow}{(\tvar{t2}, \tvar{t2})})}\\
+&\quad\quad\quad \tden{\iapp{(\ifix{f}{\darrow{\dint}{\trepof{\tvar{t2}}}}{\ilam{x}{\dint}{\\
+		\label{lastop}&\quad\quad\quad\quad\quad \iIfEq{x}{0}{\dint}{\itransof{\tvar{d2}}}{\\
+		&\quad\quad\quad\quad\quad\quad\iapp{\iapp{\itransof{\tvar{d3}}}{(x-1)}}{(\iapp{f}{(x-1)})}}}}\\
+		&\quad\quad\quad\quad)}{\itransof{\tvar{d1}}}}{\tvar{t2}}
+}}}
+}}}
+\\
+&}{XXX}{i}{\tlam{idx}{\kunit}{\titype{\dint}}};\\
+\label{nattype}&{\sf let}~\tvar{nat} = \ttype{Nat}{\tunit}~{\sf in}\\
+& \elet{two}{\eop{Nat}{s}{\tunit}{\eop{Nat}{s}{\tunit}{\eop{Nat}{z}{\tunit}{}}}}{\\
+& \elet{plus}{
+	\elam{x}{\tvar{nat}}{
+		\elam{y}{\tvar{nat}}{
+			\\&\quad \eop{Nat}{rec}{\tunit}{x; y; \elam{p}{\tvar{nat}}{\elam{r}{\tvar{nat}}{
+				\eop{Nat}{s}{\tunit}{r}
+			}}}
+		}
+	}
+}{
+& \\
+& \eop{Arrow}{ap}{\tunit}{plus; \eop{Arrow}{ap}{\tunit}{two; two}}
+}}
+%{\\&\elet{two}{\tvar{nat}}{\eop{Nat}{s}{\tunit}{\eop{Nat}{s}{\tunit}{\eop{Nat}{z}{\tunit}{ }}}}{\eapp{\eapp{plus}{two}}{two}}}
+%}
+%\elam{x}{\tvar{nat}}{\elam{y}{\tvar{nat}}{\\
+	%&\quad \eop{Nat}{rec}{\tunit}{x; y; \elam{p}{\tvar{nat}}{\elam{r}{\tvar{nat}}{
+	%\eop{Nat}{s}{\tunit}{r}
+	%}}}}
+\end{flalign}
+%\vspace{-15px}
+\caption{\small G\"odel's $\mathbf{T}$ in \atlam, used to calculate 2+2.
+The simple helper functions $\tvar{arity}n$ and $\tvar{ifeq}$ have return kind $\ksum{\kDen}{\kunit}$ (see text).
+Their definitions can be found in the appendix.
+We use \textsf{let} to bind both external and type-level variables for clarity of presentation; these can be eliminated manually.
+We will discuss the issue of a more natural syntax in Sec.
+\ref{expanded-syntax}.}
+\label{nat}
+\end{figure}
+A program, $\rho$, consists of a series of static declarations (in the core, only constructor declarations) followed by an external term, $e$.
+The syntax for external terms contains three core forms: variables, $\lambda$, and a form for all other operator invocations, which we will discuss below.
+Compiling a program consists of first \emph{kind checking} its static declarations and type-level terms (Fig.
+\ref{kindprog}, discussed below) then \emph{type checking} the external term and, simultaneously, \emph{translating} it to a term, $\iota$, in the \emph{typed internal language} (Fig.
+\ref{att}).
+The key judgement is the \emph{active typing judgement}, relating an external term to a {type} and an internal term, called its \emph{translation}, under \emph{typing context} $\Gamma$ and \emph{constructor context} $\Phi$:
 $$\ecompilesX{e}{\tau}{\iota}$$
@@
 The typing context $\Gamma$ maps variables to types and obeys standard structural properties.
-The key novelties here are that the available type and operator constructors are not fixed by the language, but rather are tracked by the constructor context, $\fvalCtx$, and that a translation is simultaneously derived.
-The dynamic semantics of external terms are defined by their translation to the internal language.
-For this judgement, the two contexts and the external term can be thought of as input, while the type and translation are output (we do not consider a bidirectional approach in this work, though this may be a promising avenue for future research).
-
-\subsection{Types}
-User-defined type constructors are declared at the top-level of a program (or package, in a practical implementation) using \textsf{tycon}.
+The key novelties here are 1) that the available type and operator constructors are not fixed by the language, but rather are tracked by the constructor context, $\fvalCtx$, and 2) that a translation is simultaneously derived.
+There is no separate operational semantics for the external language; the dynamic semantics are defined by this translation to the internal language (which might have a more conventional operational semantics).
+\todo{cite Harper-Stone / denotational semantics} The two contexts and the external term can be thought of as input to the judgement, while the type and translation are output (we do not consider a bidirectional approach \cite{LovasPfenning}\todo{cite} in this work, though this is likely a promising avenue for future research).
+
+In our calculus, the internal language (IL), consisting of internal terms $\iota$ and internal types $\sigma$, contains partial functions (via the generic fixpoint operator of Plotkin's PCF \cite{PCF}), simple product and sum types and a base type of integers (to make our example interesting and as a nod toward speed on contemporary machines).
+In practice, the internal language could be any typed  language with a specification for which type safety and decidability of typechecking have been determined (e.g.
+perhaps Standard ML or CakeML \cite{cakeml}, leaving translation to a simpler IL to later stages of the compiler).
+
+\subsection{Types and Type Constructors}
+\begin{figure}[t]
+\small
+$\fbox{\inferrule{}{\progOKX{\progsort}}}$
+%~~~$\tvarCtx ::= \emptyctx \pipe \tvarCtxX{t}{\kappa}$~~~
+%$\fCtx ::= \Sigma_0	 \pipe \fvalCtxX$
+\begin{mathpar}
+\inferrule[k-tycon]{
+	\fvar{Tycon} \notin \text{dom}(\fvalCtx)\\
+	\kEq{\kappaidx}\\
+	\tKind{\tvarCtx}{\fvalCtx}{\taurep}{\karrow{\kappaidx}{\kIType}}\\
+	\opType{\tvarCtx}{\fvalCtxX{\fvalDf}}{\theta}{\Theta}\\
+	\progOK{\tvarCtx}{\fvalCtxX{\fvalDf}}{\rho}
+}{
+	\progOKX{\pfam{\familyDf}{\rho}}
+}
+
+%\inferrule[def-kinding]{
+%	\tKindX{\tau}{\kappa}\\
+%	\progOK{\tvarCtxX{t}{\kappa}}{\fvalCtx}{\rho}
+%}{
+%	\progOKX{\pdef{t}{\kappa}{\tau}{\rho}}
+%}
+%
+\inferrule[k-e-prog]{
+	\exprOK{\tvarCtx}{\fvalCtx}{e}
+}{
+	\progOKX{e}
+}
+\end{mathpar}
+$\fbox{$\opType{\tvarCtx}{\fvalCtx}{\theta}{\Theta}$}$
+\begin{mathpar}
+\inferrule[k-opcon]{
+	\tKind{\tvarCtx}{\fvalCtx}{\taudef}{\karrow{\kappaidx}{\karrow{\klist{\kDen}}{(\ksum{\kDen}{\kunit})}}}
+}{
+	\opType{\tvarCtx}{\fvalCtx}{{\tops{op}{\kappaidx}{i}{a}{\taudef}}}{
+	\kOpS{op}{\kappaidx}}
+}
+
+\inferrule[k-opcons]{
+	\opType{\tvarCtx}{\fvalCtx}{\theta_1}{\Theta_1}\\
+	\opType{\tvarCtx}{\fvalCtx}{\theta_2}{\Theta_2}\\\\
+	\text{dom}(\theta_1) \cap \text{dom}(\theta_2) = \emptyset
+}{
+	\opType{\tvarCtx}{\fvalCtx}{\theta_1; \theta_2}{\Theta_1, \Theta_2}
+}
+\end{mathpar}
+$\fbox{$\exprOKX{e}$}$
+\begin{mathpar}
+\inferrule[k-e-var]{ }{
+	\exprOK{\tvarCtx}{\fvalCtx}{\evar{x}}
+}
+
+\inferrule[k-e-lam]{
+	\tKindX{\tau}{\kTypeBlur}\\
+	\exprOK{\tvarCtx}{\fvalCtx}{e}
+}{
+	\exprOKX{\elam{\evar{x}}{\tau}{e}}
+}
+
+\inferrule[k-e-op]{
+	\fval{Tycon}{-}{-}{\theta} \in \fvalCtx\\
+	\tops{op}{\kappaidx}{i}{a}{-} \in \theta\\
+	\tKindX{\tauidx}{\kappaidx}\\
+	\exprOKX{e_1}\\
+	\cdots\\
+	\exprOKX{e_n}
+}{
+	\exprOKX{\eop{Tycon}{op}{\tauidx}{\splat{e}{1}{n}}}
+}
+\end{mathpar}
+\caption{\small Kinding for programs.
+The kinding context $\tvarCtx$ maps type variables to kinds (we do not specify let binding of type variables at the level of programs or external terms in the core, but this would be a simple addition).
+Type-level terms stored in the constructor context are not needed during kinding, indicated using a dash.
+Kinding  for type-level terms is specified in Figure \ref{tlkind}.}
+\label{kindprog}
+\vspace{-10pt}
+\end{figure}
+\begin{figure}[t]
+\small
+$\fbox{\inferrule{}{\pcompilesX{\rho}}}$
+~~~$\fvalCtx ::= \emptyctx \pipe \fvalCtxX{\fvalDf}$
+\begin{mathpar}
+\inferrule[att-tycon]{
+	\pcompiles{\fvalCtxX{\fvalDf}}{\rho}{\gamma}
+}{
+	\pcompilesX{\pfam{\familyDf}{\rho}}
+}
+%
+%\inferrule[att-def]{
+%	\tEvalX{\tau}{\tau'}\\
+%	\pcompilesX{\subst{\tau'}{\tvar{t}}{\rho}}
+%}{
+%	\pcompilesX{\pdef{t}{\kappa}{\tau}{\rho}}
+%}
+
+\inferrule[att-e-prog]{
+	\ecompiles{\emptyctx}{\fvalCtx}{e}{\tau}{\iota}%\\
+%	\eraseX{\gabs}{\gamma}
+}{
+	\pcompilesX{e}
+}
+\end{mathpar}
+$\fbox{\inferrule{}{\ecompilesX{e}{\tau}{\iota}}}$
+~~~$\etCtx ::= \emptyctx \pipe \etCtxX{x}{\tau}$
+\begin{mathpar}
+\inferrule[att-e]{
+	\ecompilesA{\Gamma}{\fvalCtx}{e}{\tau}{\ibar}\\
+	\eraseX{\ibar}{\iota}
+}{
+	\ecompiles{\Gamma}{\fvalCtx}{e}{\tau}{\iota}\\
+}
+\end{mathpar}
+$\fbox{\inferrule{}{\ecompilesAX{e}{\tau}{\ibar}}}$
+\begin{mathpar}
+\inferrule[att-var]{ }{
+	\ecompilesA{\etCtxX{x}{\tau}}{\fvalCtx}{\evar{x}}{\tau}{\evar{x}}
+}
+
+\inferrule[att-lam]{
+	\tEvalX{\tau_1}{\tau_1'}\\
+	%\delfromtau{$\Xi_0$}{\fvalCtx}{\tau_1'}{\sabs}\\\\
+	\ecompilesA{\etCtxX{x}{\tau_1'}}{\fvalCtx}{e}{\tau_2}{\ibar}
+}{
+	\ecompilesA{\etCtx}{\fvalCtx}{\elam{\evar{x}}{\tau_1}{e}}{\ttype{Arrow}{\tpair{\tau_1'}{\tau_2}}}{\ilam{\evar{x}}{\trepof{\tau_1'}}{\ibar}}
+}
+
+\inferrule[att-op]{
+	\fval{Tycon}{-}{-}{\theta} \in \fvalCtx\\
+	\tops{op}{\kappaidx}{i}{a}{\taudef} \in \theta\\
+%	\tEvalX{\tauidx}{\tau_{\text{idx}}'}\\\\
+	\ecompilesA{\etCtx}{\fvalCtx}{e_1}{\tau_1}{\ibar_1}~~~~
+	\cdots~~~~
+	\ecompilesA{\etCtx}{\fvalCtx}{e_n}{\tau_n}{\ibar_n}\\
+	\tEvalX{\taudef~\tauidx~(\tden{\ibar_1}{\tau_1}^\checkmark :: \cdots :: \tden{\ibar_n}{\tau_n}^\checkmark :: \tnil{\kDen})}{\tinl{\kunit}{\tden{\ibar}{\tau}}}\\
+	\ddbar{\fvar{Arrow},\fvar{Tycon}}{\fvalCtx}{\trepof{\tau}}{\sbar}\\
+	\eCtxTogCtx{\fvar{Arrow},\fvar{Tycon}}{\fvalCtx}{\etCtx}{\gtCtx}\\
+	\checkRC{\gtCtx}{\fvar{Arrow},\fvar{Tycon}}{\fvalCtx}{\ibar}{\sbar}
+}{
+	\ecompiles{\etCtx}{\fvalCtx}{\eop{Tycon}{op}{\tauidx}{\splat{e}{1}{n}}}{\tau}{\ibar}
+}
+\end{mathpar}
+\caption{\small Active typechecking and translation.
+Normalization judgement for type-level terms is specified in Fig.
+\ref{tleval}.
+Local concretization ($\leadsto$)  and abstract internal typechecking ($\sim$) judgements are specified in Fig.
+\ref{?}.
+Full concretization ($\lightning$) is specified in Fig.
+\ref{?}.}
+\label{att}
+\vspace{-10pt}
+\end{figure}
+\begin{figure}[t]
+\small
+$\fbox{\inferrule{}{\ddbarX{\sbar}{\sbar'}}}$
+%~~~$\fvalCtx ::= \fvalCtx_0 \pipe \fvalCtx, \fvalDf$~~~
+~~~$\Xi ::= \emptyset \pipe \Xi, \fvar{Tycon}$~~~
+%~~~$\Xi_0 := \fvar{Arrow}$
+\begin{mathpar}
+%\inferrule[pconc-int]{ }{
+%	\ddbarX{\dint}{\dint}
+%}
+\inferrule[local-conc-show-rep]{
+	\fvar{Tycon} \in \Xi\\
+	\fval{Tycon}{-}{\taurep}{-} \in \fvalCtx\\\\
+	\tEvalX{\taurep~\tauidx}{\titype{\sbar}}\\
+	\ddbarX{\sbar}{\sbar'}
+}{
+	\ddbarX{\trepof{\ttype{Tycon}{\tauidx}}}{\sbar'}
+}
+
+\inferrule[local-conc-hide-rep]{
+	\fvar{Tycon} \notin \Xi
+}{
+	\ddbarX{\trepof{\ttype{Tycon}{\tauidx}}}{\trepof{\ttype{Tycon}{\tauidx}}}
+}
+
+\inferrule[local-conc-unquote]{
+	\ddbarX{\sbar}{\sbar'}
+}{
+	\ddbarX{\dup{\titype{\sbar}}}{\sbar'}
+}
+
+%
+\inferrule[local-conc-parr]{
+	\ddbarX{\sbar_1}{\sbar_1'}\\
+	\ddbarX{\sbar_2}{\sbar_2'}
+}{
+	\ddbarX{\darrow{\sbar_1}{\sbar_2}}{\darrow{\sbar_1'}{\sbar_2'}}
+}
+%
+%\inferrule[pconc-prod]{
+%	\ddbarX{\sbar_1}{\sbar_1'}\\
+%	\ddbarX{\sbar_2}{\sbar_2'}
+%}{
+%	\ddbarX{\dpair{\sbar_1}{\sbar_2}}{\dpair{\sbar_1'}{\sbar_2'}}
+%}
+%
+
+\text{\color{gray} (remaining rules follow recursively, omitted)}
+\end{mathpar}
+$\fbox{\inferrule{}{\eCtxTogCtxX{\etCtx}{\gtCtx}}}$
+~~~$\gtCtx ::= \emptyctx \pipe \gtCtxX{x}{\sbar}$
+\begin{mathpar}
+\inferrule[abs-empty]{ }{
+	\eCtxTogCtxX{\emptyctx}{\emptyctx}
+}
+
+\inferrule[abs-ctx]{
+	\eCtxTogCtxX{\etCtx}{\gtCtx}\\
+	\ddbarX{\trepof{\tau}}{\sbar}
+}{
+	\eCtxTogCtxX{\etCtxX{x}{\tau}}{\gtCtxX{x}{\sbar}}
+}
+\end{mathpar}
+$\fbox{\inferrule{}{\checkRCX{\ibar}{\sbar}}}$
+\begin{mathpar}
+\inferrule[abs-i-var]{ }{
+	\checkRC{\gtCtxX{x}{\sbar}}{\Xi}{\fvalCtx}{\evar{x}}{\sbar}
+}
+
+\inferrule[abs-i-fix]{
+	\ddbarX{\sbar}{\sbar'}\\
+	\checkRC{\gtCtxX{x}{\sbar'}}{\Xi}{\fvalCtx}{\ibar}{\sbar'}
+}{
+	\checkRCX{\ifix{x}{\sbar}{\gabs}}{\sbar'}
+}
+%
+%\inferrule[abs-i-lam]{
+%	\ddbarX{\sbar_1}{\sbar_1'}\\\\
+%	\checkRC{\gtCtxX{x}{\sbar_1'}}{\Xi}{\fvalCtx}{\ibar}{\sbar_2}
+%}{
+%	\checkRCX{\ilam{x}{\sbar_1}{\ibar}}{\darrow{\sbar_1'}{\sbar_2}}
+%}
+%
+%\inferrule[abs-i-ap]{
+%	\checkRCX{\ibar_1}{\darrow{\sbar_1}{\sbar_2}}\\\\
+%	\checkRCX{\ibar_2}{\sbar_1}
+%}{
+%	\checkRCX{\iapp{\ibar_1}{\ibar_2}}{\sbar_2}
+%}
+
+\text{\color{gray} (remaining rules follow from a standard statics  similarly, omitted)}
+
+%\inferrule[abs int]{ }{
+%	\checkRCX{\iintlit}{\dint}
+%}
+%
+%\inferrule[abs op]{
+%	\checkRCX{\ibar_1}{\dint}\\
+%	\checkRCX{\ibar_2}{\dint}
+%}{
+%	\checkRCX{\iop{\ibar_1}{\ibar_2}}{\dint}
+%}
+%
+%\inferrule[abs pair]{
+%	\checkRCX{\ibar_1}{\sbar_1}\\
+%	\checkRCX{\ibar_2}{\sbar_2}
+%}{
+%	\checkRCX{\ipair{\ibar_1}{\ibar_2}}{\dpair{\sbar_1}{\sbar_2}}
+%}
+%
+%\inferrule[abs fst]{
+%	\checkRCX{\ibar}{\dpair{\sbar_1}{\sbar_2}}
+%}{
+%	\checkRCX{\ifst{\ibar}}{\sbar_1}
+%}
+%
+%\inferrule[abs snd]{
+%	\checkRCX{\ibar}{\dpair{\sbar_1}{\sbar_2}}
+%}{
+%	\checkRCX{\isnd{\ibar}}{\sbar_2}
+%}
+%
+% \inferrule[abs-int-eq]{
+% 	\checkRCX{\ibar_1}{\dint}\\
+% 	\checkRCX{\ibar_2}{\dint}\\\\
+% 	\checkRCX{\ibar_3}{\sbar}\\
+% 	\checkRCX{\ibar_4}{\sbar}
+% }{
+% 	\checkRCX{\iIfEq{\ibar_1}{\ibar_2}{\dint}{\ibar_3}{\ibar_4}}{\sbar}
+% }
+%\inferrule[abs-iterm-inverse]{
+%	\checkRCX{\ibar}{\sbar}
+%}{
+%	\checkRCX{\iup{\titerm{\ibar}}}{\sbar}
+%}
+%
+\inferrule[show-trans]{
+	\fvar{Tycon} \in \Xi\\
+	\checkRCX{\ibar}{\sbar}
+}{
+	\checkRCX{\itransof{\tden{\ibar}{\ttype{Tycon}{\tauidx}}}}{\sbar}
+}
+
+\inferrule[hide-trans]{
+	\fvar{Tycon} \notin \Xi
+}{
+	\checkRCX{\itransof{\tden{\ibar}{\ttype{Tycon}{\tauidx}}^{\checkmark}}}{\trepof{\ttype{Tycon}{\tauidx}}}
+}
+\end{mathpar}
+\caption{\small Abstracted internal typing}
+\label{ait}
+\end{figure}
+@$\lambda$ is an ``actively typed'' version of the simply-typed lambda calculus with simply-kinded type-level computation.
+More specifically, type-level terms, $\tau$, themselves form a typed lambda calculus.
+The classifiers of terms at this level are called \emph{kinds}, $\kappa$ to distinguish them from  \emph{types}, which are just one kind  of type-level value.
+We will write the kind of types as $\kTypeBlur$, though it is also written $\star$, \verb|T| or \verb|Type| in various similarly structured languages (see Sec.
+\ref{related-work}).
+The kinding rules for type-level terms are given in Fig.
+\ref{todo}\todo{refer}.% Rather than there being a fixed set of type constructors, we allow the programmer to declare new type  constructors, and give the static and dynamic semantics of their associated operators, by writing type-level functions.
+In the semantics for this calculus, our kind system combined with techniques borrowed from the typed compilation literature and a form of type abstraction allow us to prove strong type safety, decidability and conservativity theorems.
+
+In most languages, types are formed by applying one of a fixed collection of \emph{type constructors} to zero or more \emph{indices}.
+In @$\lambda$, the situation is different.
+User-defined type constructors are declared at the top-level of a program (or lifted there, in a practical implementation) using \textsf{tycon} (Fig.
+\ref{kindprog}).
 Each type constructor in the program must have a unique name, written e.g.
 \fvar{Nat}.
-We do not intrinsically consider the issue of namespacing, under the assumption that globally unique names can be generated by some extrinsic mechanism (e.g.
+We assume naming conflicts can be avoided by some extrinsic mechanism (e.g.
 a URI-based scheme).
 A type constructor must also declare an \emph{index kind}, $\kappaidx$.
-Types themselves are type-level values of kind $\kTypeBlur$ and are introduced by applying a previously declared type constructor to a type-level term of its index kind, $\ttype{Tycon}{\tauidx}$.
+Types themselves -- type-level values of kind $\kTypeBlur$ -- are introduced by applying a previously declared type constructor to a type-level term of its index kind, $\ttype{Tycon}{\tauidx}$.
 The kind $\kTypeBlur$ also has an elimination form: a type can be case analyzed against a type constructor in scope to extract its index.
-Like open datatypes, there is no longer a notion of exhaustiveness.
-To maintain totality, we require the default case.
-
-To permit the implementation of interesting type systems, the type-level language includes several other kinds of data alongside types.
+We are, to a first approximation, treating type constructors as constructors of a built-in open datatype at the type-level.
+Like open datatypes, there is no longer a notion of exhaustiveness and the default case is required for case analysis.
+
+ To permit the implementation of interesting type systems, the type-level language includes several kinds other than $\kTypeBlur$.
 We lift several  standard functional constructs to the type level: unit ($\kunit$), binary sums ($\ksum{\kappa_1}{\kappa_2}$), binary products ($\kpair{\kappa_1}{\kappa_2}$), lists ($\klist{\kappa}$) and integers ($\dint$).
 We also include labels ($\klabel$), written in a slanted font, e.g.
-$\tlabel{myLabel}$, which are atomic string-like values that can only be compared, here only for equality, and play a distinguished role in the expanded syntax, as we will later discuss.
+$\tlabel{myLabel}$, which are string-like values that only admit comparisons, here only for equality, and play a distinguished role in the expanded syntax, as we will later discuss.
 Our first example, $\fvar{Nat}$, is indexed trivially, i.e.
 by unit kind, $\kunit$, because there is only one natural number type, $\ttype{Nat}{\tunit}$, but we will show examples of type constructors that are indexed in more interesting ways in later portions of this work.
-For example, $\fvar{Tuple}$ is indexed by a list of types and $\fvar{LabeledTuple}$ is indexed by a list pairing labels with types.
+For example, $\fvar{Tuple}$ has index kind $\klist{\kTypeBlur}$ and $\fvar{LabeledTuple}$ has index kind $\klist{\kpair{\klabel}{\kTypeBlur}}$.
+The type constructor $\fvar{Arrow}$ is included in the initial constructor context and has index kind $\kpair{\kTypeBlur}{\kTypeBlur}$.
 
 Two type-level terms of kind $\kTypeBlur$ are equivalent if they apply the same constructor, identified by name, to equivalent indices.
-Going further, we ensure that deciding type equivalence requires only checking for syntactic equality after evaluation to normal form by imposing the restriction that type constructors can only be indexed by kinds for which equivalence can be decided in this way.
+Going further, we ensure that deciding type equivalence requires only checking for syntactic equality after normalization by imposing the restriction that type constructors can only be indexed by kinds for which equivalence can be decided in this way.
 All combinations of kinds introduced thus far  have this property and this is sufficient for many interesting examples.
-We leave  introducing a richer notion of  equivalence of types that  extension providers can extend as future work, as the metatheoretic guarantee that typing respects type equivalence would be quite a bit more complex in such a setting.% (a na\"ive approach to this would impose non-trivial extrinsic proof obligations onto extension developers that, unlike in others in this thesis, could threaten type safety).
+We discuss introducing a richer notion of  equivalence of types that  fragment providers can extend as future work in Sec.
+\ref{limitations}\todo{do this}.% as maintaining the metatheoretic guarantee that typing respects type equivalence would impose a substantial burden in such a setting.% (a na\"ive approach to this would impose non-trivial extrinsic proof obligations onto extension developers that, unlike in others in this thesis, could threaten type safety).
 
@@
 22 of \emph{PFPL} describes a related system \cite{pfpl}).
-The type-level language does, however, include total functions of conventional arrow kind, $\karrow{\kappa_1}{\kappa_2}$.
-Type constructor application can be wrapped in a type-level function to emulate first-class type constructors (and indeed, such a wrapper could be generated automatically, though we avoid this for simplicity in our semantics).
+The type-level language does, however, include total functions of conventional arrow kind, written $\karrow{\kappa_1}{\kappa_2}$.
+Type formation can be wrapped in a type-level function to emulate a first-class type constructor (indeed, such a wrapper could be generated automatically, though we avoid this for simplicity in our semantics).
 Equivalence at arrow kind does not coincide with $\beta$-equivalence, so type-level functions cannot appear in type indices.
 Our treatment of equivalence in the type-level language is thus quite similar to the treatment of term-level equality using ``equality types'' in a language like Standard ML.
-Indeed, one might imagine that a practical implementation of our calculus would start with a pure, total subset of the term language of ML to construct the type-level language (consistent with its early development as a ``metalanguage'' and, as we will see, the role of our type-level language as a DSL for implementing type systems and translators, which functional languages are widely considered well-suited for as-is).
-We leave the development of such an implementation (and of bootstrapping type-level and internal languages that are themselves user-defined or extensible) as areas for future work.
+An equality type is defined by the judgement \todo{eq type discussion}.
+Every type constructor also defines a \emph{representation schema} using \textsf{schema}, a type-level function that we will discuss after introducing operators.
+
 \subsection{Operators}
-User-defined operator constructors are declared using \textsf{opcon}.
-For reasons that we will discuss, our calculus associates every operator constructor with a type constructor.
-The \emph{fully-qualified name} of the operator constructor, e.g.
+For reasons that we will discuss, our calculus associates every operator  constructor\footnote{We distinguish \emph{term constructors} from \emph{type constructors} and \emph{operator constructors}.
+There are term constructors at all levels in the calculus.
+For example, type constructor application is the sole type-forming type-level term constructor.
+An abstract syntax at the type level like the one used by Harper \cite{pfpl} might write it $\texttt{tcapp}[\fvar{Tycon}](\tau)$, emphasizing this point.} with a type constructor.
+
+User-defined operator constructors are declared using \textsf{opcon} (Fig.
+\ref{kindprog}).
+The \emph{fully-qualified name} of every operator constructor, e.g.
 $\fvar{Nat}.\opvar{z}$, must be unique.
 Operator constructors are indexed by type-level values, like type constructors, and so also declare an index kind $\kappaidx$.
-In our first example, all the operator constructors are indexed trivially, but later examples will use more interesting indices.
-For example, the projection operators for tuples and labeled tuples use numbers and labels as indices, respectively.
-Note that neither operator constructors nor operators are first-class type-level values (we plan to investigate  lifting the latter restriction) and we do not impose any restrictions on their index kinds.
-
-In the external language, an operator is selected and invoked by applying an operator constructor to a type-level index and $n \geq 0$ \emph{arguments}, $\eop{Tycon}{op}{\tauidx}{\splat{e}{1}{n}}$.
+In our first example, all the operator constructors are indexed trivially (by index kind $\kunit$), but later examples will use more interesting indices.
+For example, the projection operator for labeled tuples, $\fvar{LabeledTuple}.\opvar{prj}$, has index kind $\klabel$.
+Note that neither operator constructors nor operators are first-class type-level values and we do not need to impose any restrictions on their index kinds.
+
+In the external language, an operator is selected and invoked by applying an operator constructor to a type-level index and $n \geq 0$ \emph{arguments}, written $\eop{Tycon}{op}{\tauidx}{\splat{e}{1}{n}}$.
 For example, on line 18 of Fig.
-\ref{nat}, we see the operator constructors $\fvar{Nat}.\opvar{z}$ and $\fvar{Nat}.\opvar{s}$ being invoked to compute two.\footnote{Although our focus here is entirely on semantics, a brief note on syntax: in the expanded syntax, the trivial indices and empty argument lists can be omitted, so we could write \texttt{Nat.s(Nat.s(Nat.z))}.
+\ref{nat}, we see the operator constructors $\fvar{Nat}.\opvar{z}$ and $\fvar{Nat}.\opvar{s}$ being invoked to compute two (the syntactic overhead will be relieved later).
+%\footnote{Although our focus here is entirely on semantics, a brief note on syntax: in the expanded syntax, the trivial indices and empty argument lists can be omitted, so we could write \texttt{Nat.s(Nat.s(Nat.z))}.
 With the ability to ``open'' a type's operators into the context, we could shorten this still to \texttt{s(s(z))}.
@@
 \subsection{Representational Consistency Implies Type Safety}
+
 In our example, natural numbers are represented internally as integers.
@@
 For example, the following operator declaration in some other type that we have loaded would topple our house of cards\todo{TBC}:
+
+\section{Semantics and Metatheory}\label{theory}
+
+\section{Expanded Syntax}\label{expanded-syntax}
+
+\section{Examples}\label{examples}
+
+\section{Limitations}\label{limitations}
+\todo{integrate this}Indeed, it is not a stretch to imagine a reasonably faithful implementation of our calculus starting with a pure, total subset of the term language of ML to construct the type-level language.
+We leave the development of such an implementation (and of bootstrapping type-level and internal languages that are themselves user-defined or extensible) as areas for future work to focus sharply on the theoretical foundations in this work.
 
@@
 %\inferrule{
-%	\overbrace{\progOK{\emptyctx}{\fvarCtx_0}{\rho}}^{\text{\normalsize Kind Checking}}\\
+%	\overbrace{\progOK{\emptyctx}{\fvalCtx_0}{\rho}}^{\text{\normalsize Kind Checking}}\\
 %	\overbrace{\pcompiles{\fvalCtx_0}{\rho}{\gamma}}^{\text{\normalsize Active Typechecking and Translation}}
@@
 %~~~$\tvarCtx ::= \emptyctx \pipe \tvarCtxX{t}{\kappa}$~~~
-%$\fCtx ::= \Sigma_0	 \pipe \fvarCtxX$
+%$\fCtx ::= \Sigma_0	 \pipe \fvalCtxX$
 %\begin{mathpar}
 %\inferrule[family-kinding]{
-%	\fvar{Fam} \notin \text{dom}(\fvarCtx)\\
+%	\fvar{Fam} \notin \text{dom}(\fvalCtx)\\
 %	\kEq{\kappaidx}\\
-%	\opType{\tvarCtx}{\fvarCtxX}{\theta}{\Theta}\\\\
-%	\tKind{\tvarCtx, \tvar{i}:{\kappaidx}}{\fvarCtxX}{\tau}{\kIType}\\
-%	\progOK{\tvarCtx}{\fvarCtxX}{\rho}
+%	\opType{\tvarCtx}{\fvalCtxX}{\theta}{\Theta}\\\\
+%	\tKind{\tvarCtx, \tvar{i}:{\kappaidx}}{\fvalCtxX}{\tau}{\kIType}\\
+%	\progOK{\tvarCtx}{\fvalCtxX}{\rho}
 %}{
@@
 %	\tKindX{\tau}{\kappa}\\
-%	\progOK{\tvarCtxX{t}{\kappa}}{\fvarCtx}{\rho}
+%	\progOK{\tvarCtxX{t}{\kappa}}{\fvalCtx}{\rho}
 %}{
@@
 %\inferrule[exp-kinding]{
-%	\exprOK{\tvarCtx}{\emptyctx}{\fvarCtx}{e}
+%	\exprOK{\tvarCtx}{\emptyctx}{\fvalCtx}{e}
 %}{
@@
 %\inferrule[op-kinding]{
-%	\tKind{\tvarCtx, \tOfKind{\tvar{i}}{\kappai}, \tOfKind{\tvar{a}}{\klist{\kDen}}}{\fvarCtx}{\tau}{\kDen}
-%}{
-%	\opType{\tvarCtx}{\fvarCtx}{{\tops{op}{\kappaidx}{i}{a}{\tau}}}{
+%	\tKind{\tvarCtx, \tOfKind{\tvar{i}}{\kappai}, \tOfKind{\tvar{a}}{\klist{\kDen}}}{\fvalCtx}{\tau}{\kDen}
+%}{
+%	\opType{\tvarCtx}{\fvalCtx}{{\tops{op}{\kappaidx}{i}{a}{\tau}}}{
 %	\kOpS{op}{\kappaidx}}
@@
 %\inferrule[e-var-kinding]{ }{
-%	\exprOK{\tvarCtx}{\itvarCtx, \evar{x}}{\fvarCtx}{\evar{x}}
+%	\exprOK{\tvarCtx}{\itvarCtx, \evar{x}}{\fvalCtx}{\evar{x}}
 %}
@@
 %	\tKindX{\tau}{\kTypeBlur}\\
-%	\exprOK{\tvarCtx}{\eivarCtxX{x}}{\fvarCtx}{e}
+%	\exprOK{\tvarCtx}{\eivarCtxX{x}}{\fvalCtx}{e}
 %}{
@@
 %\inferrule[e-op-kinding]{
-%	\fvarOfType{Fam}{\kappaidx}{\Theta} \in \fvarCtx\\
+%	\fvarOfType{Fam}{\kappaidx}{\Theta} \in \fvalCtx\\
 %	\kOpS{op}{\kappai} \in \Theta\\
@@
 %%\subsection{Active Typechecking and Translation}
-%\begin{figure}[t]
-%\small
-%$\fbox{\inferrule{}{\pcompilesX{\rho}}}$
-%~~~$\fvalCtx ::= \fvalCtx_0 \pipe \fvalCtxX{\fvalDf}$
-%\begin{mathpar}
-%\inferrule[att-fam]{
-%	\pcompiles{\fvalCtxX{\fvalDf}}{\rho}{\gamma}
-%}{
-%	\pcompilesX{\pfam{\familyDf}{\rho}}
-%}
-%
-%\inferrule[att-def]{
-%	\tEvalX{\tau}{\tau'}\\
-%	\pcompilesX{\subst{\tau'}{\tvar{t}}{\rho}}
-%}{
-%	\pcompilesX{\pdef{t}{\kappa}{\tau}{\rho}}
-%}
-%
-%\inferrule[att-exp]{
-%	\ecompiles{\emptyctx}{\fvalCtx}{e}{\tau}{\gabs}\\
-%	\eraseX{\gabs}{\gamma}
-%}{
-%	\pcompilesX{e}
-%}
-%\end{mathpar}
-%$\fbox{\inferrule{}{\ecompilesX{e}{\tau}{\gabs}}}$
-%~~~$\etCtx ::= \emptyctx \pipe \etCtxX{x}{\tau}$
-%\begin{mathpar}
-%\inferrule[att-var]{ }{
-%	\ecompiles{\etCtxX{x}{\tau}}{\fvalCtx}{\evar{x}}{\tau}{\evar{x}}
-%}
-%
-%\inferrule[att-lam]{
-%	\tEvalX{\tau_1}{\tau_1'}\\
-%	%\delfromtau{$\Xi_0$}{\fvalCtx}{\tau_1'}{\sabs}\\\\
-%	\ecompiles{\etCtxX{x}{\tau_1'}}{\fvalCtx}{e}{\tau_2}{\gabs}
-%}{
-%	\ecompiles{\etCtx}{\fvalCtx}{\elam{\evar{x}}{\tau_1}{e}}{\ttype{Arrow}{\tpair{\tau_1'}{\tau_2}}}{\ilam{\evar{x}}{\trepof{\tau_1'}}{\gabs}}
-%}
-%
-%\inferrule[att-op]{
-%	\fvalDf \in \fvalCtx\\
-%	\tops{op}{\kappaidx}{i}{a}{\tauop} \in \theta\\
-%	\tEvalX{\tauidx}{\tauidx'}\\\\
-%	\ecompiles{\etCtx}{\fvalCtx}{e_1}{\tau_1}{\gamma_1}~~~~
-%	\cdots~~~~
-%	\ecompiles{\etCtx}{\fvalCtx}{e_n}{\tau_n}{\gamma_n}\\
-%	\text{$\begin{array}{c}
-%		\left[\text{$\begin{array}{r}
-%			\tauidx'/\tvar{i}\\
-%			\tden{\titerm{\gamma_1}}{\tau_1} :: \cdots :: \tden{\titerm{\gamma_n}}{\tau_n} :: \tnil{\kDen}/\tvar{a}
-%			\end{array}$}\right]\tauop
-%		\Downarrow
-%		\tden{\titerm{\gabs}}{\tau}
-%		\end{array}$}\\
-%	%\delfromtau{Fam}{\fvalCtx}{\tau}{\sabs}\\
-%	%
-%	\ddbar{\Xi_0,\fvar{Fam}}{\fvalCtx}{\trepof{\tau}}{\sabs}\\
-%	\eCtxTogCtx{\Xi_0,\fvar{Fam}}{\fvalCtx}{\etCtx}{\gtCtx}\\
-%	\checkRC{\gtCtx}{\Xi_0,\fvar{Fam}}{\fvalCtx}{\gabs}{\sabs}
-%}{
-%	\ecompiles{\etCtx}{\fvalCtx}{\eop{Fam}{op}{\tauidx}{\splat{e}{1}{n}}}{\tau}{\gabs}
-%}
-%\end{mathpar}
-%\caption{\small Active typechecking and translation of programs. Evaluation semantics for type-level terms are given in Fig. \ref{tleval}.}
-%\label{att}
-%\vspace{-10pt}
-%\end{figure}
 %The second phase of compilation involves invoking the  logic implemented by user-defined operator families to typecheck and translate the program.
@@
 %}{
-%  \tKind{\tvarCtxX{t}{\kappa}}{\fvarCtx}{\tvar{t}}{\kappa}
+%  \tKind{\tvarCtxX{t}{\kappa}}{\fvalCtx}{\tvar{t}}{\kappa}
 %}
@@
 %\inferrule[k-arrow-intro]{
-%  \tKind{\tvarCtxX{t}{\kappa_1}}{\fvarCtx}{\tau}{\kappa_2}
+%  \tKind{\tvarCtxX{t}{\kappa_1}}{\fvalCtx}{\tau}{\kappa_2}
 %}{
@@
 %\inferrule[type-intro]{
-%	\fvarOfTypeDf \in \fvarCtx\\\\
-%	\tKind{\tvarCtx}{\fvarCtx}{\tauidx}{\kappaidx}
+%	\fvarOfTypeDf \in \fvalCtx\\\\
+%	\tKind{\tvarCtx}{\fvalCtx}{\tauidx}{\kappaidx}
 %}{
@@
 %\inferrule[type-elim]{
-%	\fvarOfTypeDf \in \fvarCtx\\
+%	\fvarOfTypeDf \in \fvalCtx\\
 %	\tKindX{\tau}{\kTypeBlur}\\\\
-%	\tKind{\tvarCtxX{x}{\kappaidx}}{\fvarCtx}{\tau_0}{\kappa}\\
+%	\tKind{\tvarCtxX{x}{\kappaidx}}{\fvalCtx}{\tau_0}{\kappa}\\
 %	\tKindX{\tau_1}{\kappa}
 %}{
-%	\tKind{\tvarCtx}{\fvarCtx}{\tfamcase{\tau}{Fam}{x}{\tau_0}{\tau_1}}{\kappa}
+%	\tKind{\tvarCtx}{\fvalCtx}{\tfamcase{\tau}{Fam}{x}{\tau_0}{\tau_1}}{\kappa}
 %}
@@
 %	\tKindX{\tau}{\kDen}\\\\
-%	\tKind{\tvarCtx, \tOfKind{\tvar{x}}{\kITerm}, \tOfKind{\tvar{t}}{\kTypeBlur}}{\fvarCtx}{\tau_1}{\kappa}\\
+%	\tKind{\tvarCtx, \tOfKind{\tvar{x}}{\kITerm}, \tOfKind{\tvar{t}}{\kTypeBlur}}{\fvalCtx}{\tau_1}{\kappa}\\
 %	\tKindX{\tau_2}{\kappa}
@@
 %\inferrule[iterm-intro]{
-%	\isIterm{\tvarCtx}{\emptyctx}{\fvarCtx}{\gamma}
+%	\isIterm{\tvarCtx}{\emptyctx}{\fvalCtx}{\gamma}
 %}{
@@
 %\inferrule[itype intro]{
-%	\isItype{\tvarCtx}{\fvarCtx}{\sigma}
+%	\isItype{\tvarCtx}{\fvalCtx}{\sigma}
 %}{
@@
 %\inferrule[i-var-kinding]{ }{
-%	\isIterm{\tvarCtx}{\eivarCtxX{x}}{\fvarCtx}{\evar{x}}
+%	\isIterm{\tvarCtx}{\eivarCtxX{x}}{\fvalCtx}{\evar{x}}
 %}
@@
 %	\isItypeX{\sigma}\\\\
-%	\isIterm{\tvarCtx}{\eivarCtxX{x}}{\fvarCtx}{\gamma}
+%	\isIterm{\tvarCtx}{\eivarCtxX{x}}{\fvalCtx}{\gamma}
 %}{
@@
 %
-%\section{Related Work}\label{related-work}
-%Our representation schema abstraction mechanism relates closely to abstract and existential types \cite{pfpl,atpl}.
+\section{Related Work}\label{related-work}
+Our representation schema abstraction mechanism relates closely to abstract and existential types \cite{pfpl,atpl}.
 Our calculus enforces the abstraction barriers  in a purely syntactic manner, as in previous work on syntactic type abstraction \cite{syntypeabs}.
@@
 a module), we focus on abstracting away the knowledge of how a primitive type family is implemented outside of a limited scope.
-%
-%The representational consistency mechanism brings into the language work on typed compilation, especially work done on Standard ML in the TILT project \cite{tilt}.
+
+The representational consistency mechanism brings into the language work on typed compilation, especially work done on Standard ML in the TILT project \cite{tilt}.
 Indeed, the specification of Standard ML is structured around a typed internal language and a judgement that assigns a type and an internal term to each expression \cite{smlstd}.
 Representational consistency is related to the notion of type-directed copmilation in this work.
-%
-%Type-level computation is supported in some form by a growing number of languages.
+
+Type-level computation is supported in some form by a growing number of languages.
 For example, Haskell supports a simple form of it \cite{Chakravarty:2005:ATC}).
@@
 We show how to integrate language extensions into the type-level language, drawing on ideas about \cite{activelibraries}.
-%
-%% \subsection{Run-Time Indirection}
-%% {\it Operator overloading} \cite{vanWijngaarden:Mailloux:Peck:Koster:Sintzoff:Lindsey:Meertens:Fisker:acta:1975} and {\it metaobject dispatch} \cite{Kiczales91} are run-time protocols that translate operator invocations into function calls.
+
+% \subsection{Run-Time Indirection}
+% {\it Operator overloading} \cite{vanWijngaarden:Mailloux:Peck:Koster:Sintzoff:Lindsey:Meertens:Fisker:acta:1975} and {\it metaobject dispatch} \cite{Kiczales91} are run-time protocols that translate operator invocations into function calls.
 The function is typically selected according to the type or value of one or more operands.
@@
 However, type-level specification is a {\it compile-time} protocol focused on enabling specialized verification and implementation strategies, rather than simply enabling run-time indirection.
-%
-%Many languages and tools allow developers to rewrite expressions according to custom rules.
+
+Many languages and tools allow developers to rewrite expressions according to custom rules.
 These can broadly be classified as {\it term rewriting systems}.
@@
 MetaML \cite{Sheard:1999:UMS}), and external rewrite systems also exist for many languages.
-%These facilities differ from AT\&T in that they involve direct manipulation of terms, while AT\&T involves extending typechecking and translation logic directly.
+These facilities differ from AT\&T in that they involve direct manipulation of terms, while AT\&T involves extending typechecking and translation logic directly.
 We also draw a distinction between the type-level, used to specify types and compile-time logic, the expression grammar, used to describe run-time behavior, and the internal language, used to implement this behavior.
 By doing so, each component can be structured and constrained as appropriate for its distinct role, as we have shown.
-%
-%Previous work on extensible languages has suffered from problems with either expressiveness or safety.
+
+Previous work on extensible languages has suffered from problems with either expressiveness or safety.
 For example, a number of projects, such as SugarJ \cite{erdweg2011sugarj}, allow for user-defined desugarings (and indeed, our system would clearly benefit from integration with such a mechanism), but this does not allow the semantics to be fundamentally extended nor for implementation details to be hidden.
@@
 Our work aims to provide a sound and safe underpinning, based around well-understood concepts of type and operator families, to language extensibility.
-%
-%In the future, we will investigate mechanisms to enable type systems with specialized binding and scoping rules, as well as integration of dependent kinds to support mechanized verification of key properties like representational consistency and adequacy against a declarative specification at kind-checking time.
+
+In the future, we will investigate mechanisms to enable type systems with specialized binding and scoping rules, as well as integration of dependent kinds to support mechanized verification of key properties like representational consistency and adequacy against a declarative specification at kind-checking time.
 We will also investigate mechanisms that enable a more natural syntax (e.g.
 Wyvern's type-directed syntax \cite{globaldsl}).
-%
-%%\item With type-level specification, dispatch to a type-level function occurs implicitly on the basis of the structure of an expression.
+
+%\item With type-level specification, dispatch to a type-level function occurs implicitly on the basis of the structure of an expression.
 In contrast, most term-rewriting systems operate by  explicit invocation of a macro or specialized syntax.
@@
 A*\cite{fowler2010domain}, EPP\cite{fowler2010domain}) and macro systems for object-oriented languages, like OpenC++ \cite{fowler2010domain} and OpenJava \cite{fowler2010domain}, do offer a somewhat limited form of operation-based dispatch.
-%
-%% \subsection{Language Frameworks}
-%% When the mechanisms available in an existing language prove insufficient, researchers and domain experts must design a new language.
+
+% \subsection{Language Frameworks}
+% When the mechanisms available in an existing language prove insufficient, researchers and domain experts must design a new language.
 A number of tools have been developed to assist with this task, including compiler generators, language workbenches and domain-specific language frameworks (cf \cite{fowler2010domain}).
-%
-%% A major barrier to adoption is the fact that interoperability is intrinsically problematic.
+
+% A major barrier to adoption is the fact that interoperability is intrinsically problematic.
 Even languages which target a common platform, such as the Java Virtual Machine, can only interact using its limited set of primitives.
@@
 Java).
-%
-%% Instead of focusing on defining standalone languages, type-level specification gives greater responsibility in a granular manner to libraries.
+
+% Instead of focusing on defining standalone languages, type-level specification gives greater responsibility in a granular manner to libraries.
 In this way, a range of constructs can coexist within the same program and, assuming that it can be shown by some method that various constructs are safely composable, be mixed and matched.
@@
 Nevertheless, as Ace shows, a simple grammar can be used quite flexibly.
-%% \subsection{Extensible Compilers}
-%% An alternative methodology is to implement language features granularly as compiler extensions.
+% \subsection{Extensible Compilers}
+% An alternative methodology is to implement language features granularly as compiler extensions.
 As discussed in Section 1, existing designs suffer from the same problems related to composability, modularity\-, safety and security as extensible languages, while also adding the issue of language fragmentation.
-%
-%% Type-level specification can in fact be implemented within a compiler, rather than provided as a core language feature.
+
+% Type-level specification can in fact be implemented within a compiler, rather than provided as a core language feature.
 This would resolve some of the issues, as described in this paper.
@@
 We also avoid the fragmentation issue.
-%% \subsection{Specification Languages}
-%% Several {\it specification languages} (or {\it logical frameworks}) based on these theoretical formulations exist, including the OBJ family of languages (e.g.
+% \subsection{Specification Languages}
+% Several {\it specification languages} (or {\it logical frameworks}) based on these theoretical formulations exist, including the OBJ family of languages (e.g.
 CafeOBJ \cite{Diaconescu-Futatsugi01}).
@@
 The  language itself specifies which verification and execution strategies are used.
-%
-%% Type-determined compilation takes a more concrete approach to the problem, focusing on combining {\it implementations} of different\- logics, rather than simply their specifications.
+
+% Type-determined compilation takes a more concrete approach to the problem, focusing on combining {\it implementations} of different\- logics, rather than simply their specifications.
 In other words, it focuses on combining {\it type checkers} and {\it implementation strategies} rather than more abstract representations of a language's type system and dynamic semantics.

### 2014-02-28 att-icfp14/att-icfp14.tex
More icfp progress

@@
 
-\title{@$\lambda$: Conservatively Extending a Type System From Within}
+\title{Active Typing: Conservatively Extending a Type System From Within}
 
@@
 %Having established why a minimal language like $\mathcal{L}\{\rightarrow\forall\}$, while suitable as an internal language, needs to be extended with additional fragments before it is suitable for use as a human-facing external language,
-``General-purpose'' languages like ML, on the other hand, expose mechanisms that admit strong embeddings of many more fragments and occupy what is widely seen as a ``sweet spot'' in the design space.
-Nevertheless, situations continue to arise in both research and practice where introducing a new fragment would be desirable because an abstraction can still only be weakly defined in terms of existing fragments, or a strong embedding, while possible, is impractically verbose or inefficient:
+Modern ``general-purpose'' languages like ML, on the other hand, expose constructs, like datatypes and abstract types, that admit strong embeddings of many simpler fragments, occupying what is widely seen as a ``sweet spot'' in the design space.
+However, ``general-purpose'' and ''all-purpose'' are quite distinct, and situations continue to arise in both research and practice where introducing a new fragment would be desirable because it can still only be weakly defined in terms of existing fragments, or a strong embedding, while possible, is impractically verbose or inefficient:
 \begin{enumerate}
@@
 Forming such a dialect is alarmingly non-trivial, even in the rare situation where a common framework has been used or the dialects are implemented within the same compiler, as these mechanisms do not guarantee that different combinations of individually sound language fragments remain sound when combined.
-Metatheoretic and compiler correctness results can only be derived for the dialect \emph{resulting} from a language   composition operation, so fragment providers have little choice but to leave this to clients.% We will consider this further in our discussion of related work in Sec.
+Metatheoretic and compiler correctness results can only be derived for the dialect \emph{resulting} from a language   composition operation, so in a diverse fragment ecosystem, fragment providers have little choice but to leave this to clients.
+Avoiding language composition is difficult, as interoperability between components of an application written in different dialects faces precisely the problems described in item 3.% We will consider this further in our discussion of related work in Sec.
 \ref{related-work}.
@@
 \paragraph{Contributions}
-We draw inspiration from this approach to design a mechanism that makes it possible to declare new type and operator constructors and define,  {using type-level functions}, their static and dynamic semantics, the latter by translation to a fixed typed internal language, from within.
-If a language fragment can be weakly defined in terms of the typed internal language and its statics are of a form that can be implemented by this protocol, it can be made strongly embeddable in the external language by introducing, in this way, the necessary constructors and logic.
-
-The protocol constrains extensions to maintain type safety and decidability of type checking and type equality.
+We draw inspiration from this approach to design a mechanism that makes it possible to declare new type and operator constructors and define, \emph{using type-level functions}, their static and dynamic semantics, the latter by translation to a fixed typed internal language.
+If a language fragment can be weakly defined in terms of the typed internal language and its statics are of a form that can be implemented by this protocol, it can be made strongly embeddable in the external language by introducing, in this way, the necessary constructors and static logic.
+
+The semantics imposes checks on extensions to guarantee type safety and decidability of typing and type equality.
 Moreover, an important class of lemmas that play a key role in  proofs that an extension enables strong embeddings of a language fragment can be derived assuming a ``closed world'' (that is, in the completely traditional manner where one constructs a complete language and reasons inductively).
-The semantics guarantees that they will be conserved in the  ``open world'' by enforcing a form of  representation independence at all extension boundaries.
-Thus, if a strong embedding can be constructed using some combination of extensions, it can be constructed in the presence of arbitrary combinations of other extensions as well, avoiding the problems faced by language dialects and justifying the inclusion of the mechanism directly into the language.
+The semantics guarantees that they will be \emph{conserved} in the  ``open world'' by enforcing a form of  representation independence at all extension boundaries.
+This avoids the biggest problems of composing language dialects and justifies the inclusion of the mechanism directly into the language.
 
 We will begin in Sec.
-\ref{core} by introducing a minimal calculus with an actively typed semantics, @$\lambda$.
-The external language builds in only the $\rightarrow$ fragment and the typed internal language is a simple variant of PCF that permits only weak embeddings of even simple fragments like the $\mathtt{nat}$ fragment of G\"odel's $\mathbf{T}$ or ML-style $n$-ary products.
-We implement these as extensions to explain the mechanism and motivate the constraints that the semantics imposes to make the guarantees that we claim.
-We then examine the semantics in greater detail and state and describe the key points in the proofs of the metatheory in Sec.
+\ref{core} by introducing a minimal calculus with this kind of \emph{actively typed semantics}, @$\lambda$.
+The external language begins with only the $\rightarrow$ fragment and the typed internal language is a simple variant of PCF that permits only weak embeddings of even fragments like the $\mathtt{nat}$ fragment of G\"odel's $\mathbf{T}$ or ML-style $n$-ary products.
+We implement these as extensions to explain the mechanism and motivate the constraints that the semantics imposes.
+We then examine the semantics in greater detail and state and sketch the key points in the proofs of the metatheory in Sec.
 \ref{theory}.
-These two sections contain the most fundamental contributions of this work.
+These are the fundamental contributions of this work.
 
@@
 
-While of surprising expressiveness, @$\lambda$ itself is certainly not capable of expressing the full variety of type systems listed above.
-Indeed, to guarantee safety and conservativity, the calculus is perhaps too conservative.
+While of surprising expressiveness, @$\lambda$ itself is certainly not capable of expressing the full variety of type systems listed earlier.
+Indeed, to guarantee safety and conservativity, our calculus assumes perhaps too much about the type system being encoded.
 For example, it only admits fragments for which the typing judgement looks essentially like the typing judgement of the $\rightarrow$ fragment; new contexts cannot be introduced.
@@
 \ref{limitations}.
-These limitations do not appear to be fundamental to the approach, so we also outline in Sec.
-\ref{limitations} promising directions for future work.
+Fortunately, these limitations do not appear fundamental to the approach, so we also outline in Sec.
+\ref{limitations} directions for future work.
 We conclude with a discussion of related work in Sec.
@@
 \ref{nat}.
-We will discuss its semantics and how precisely the strong embedding seen being used in this example works as we go on.
+We will discuss its semantics and how precisely the strong embedding, seen being used starting on line 16, works as we go on.
 Natural numbers can, of course, be strongly embedded in existing languages with a similar usage and asymptotic performance profile (up to function call overhead with an abstract type, for example).
@@
 \end{figure}
-A program, $\rho$, consists of a series of static declarations (in the core, only constructor declarations) followed by an external term, $e$.
-The syntax for external terms contains three core forms: variables, $\lambda$, and a form for all other operator invocations, which we will discuss below.
+A \emph{program}, $\rho$, consists of a series of static declarations (in the core, only constructor declarations) followed by an external term, $e$.
+The syntax for external terms contains three term formers: variables, $\lambda$, and a form for all other operator invocations, which we will discuss below.
 Compiling a program consists of first \emph{kind checking} its static declarations and type-level terms (Fig.
-\ref{kindprog}, discussed below) then \emph{type checking} the external term and, simultaneously, \emph{translating} it to a term, $\iota$, in the \emph{typed internal language} (Fig.
+\ref{kindprog}, discussed below) then \emph{type checking} the external term and, simultaneously, \emph{translating} it to a term, $\iota$, in the {typed internal language} (Fig.
 \ref{att}).
@@
 There is no separate operational semantics for the external language; the dynamic semantics are defined by this translation to the internal language (which might have a more conventional operational semantics).
-\todo{cite Harper-Stone / denotational semantics} The two contexts and the external term can be thought of as input to the judgement, while the type and translation are output (we do not consider a bidirectional approach \cite{LovasPfenning}\todo{cite} in this work, though this is likely a promising avenue for future research).
-
-In our calculus, the internal language (IL), consisting of internal terms $\iota$ and internal types $\sigma$, contains partial functions (via the generic fixpoint operator of Plotkin's PCF \cite{PCF}), simple product and sum types and a base type of integers (to make our example interesting and as a nod toward speed on contemporary machines).
-In practice, the internal language could be any typed  language with a specification for which type safety and decidability of typechecking have been determined (e.g.
-perhaps Standard ML or CakeML \cite{cakeml}, leaving translation to a simpler IL to later stages of the compiler).
+\todo{cite Harper-Stone / denotational semantics} The two contexts and the external term can be thought of as input to the judgement, while the type and translation are output (we do not consider a bidirectional approach \cite{LovasPfenning}\todo{cite} or type inference in this work, though this is likely a promising avenue for future research).
+
+In our calculus, the internal language (IL), consisting of internal terms $\iota$ and internal types $\sigma$, provides partial functions (via the generic fixpoint operator of Plotkin's PCF \cite{PCF}), simple product and sum types and a base type of integers (to make our example interesting and as a nod toward speed on contemporary machines).
+In practice, the internal language could be any typed  language with a specification for which type safety and decidability of typechecking have been satisfyingly determined (e.g.
+Standard ML or CakeML \cite{cakeml}, leaving translation to a simpler IL to later stages of the compilation process).
 
@@
 @$\lambda$ is an ``actively typed'' version of the simply-typed lambda calculus with simply-kinded type-level computation.
-More specifically, type-level terms, $\tau$, themselves form a typed lambda calculus.
-The classifiers of terms at this level are called \emph{kinds}, $\kappa$ to distinguish them from  \emph{types}, which are just one kind  of type-level value.
+Specifically, type-level terms, $\tau$, themselves form a typed lambda calculus.
+The classifiers of terms at the type level are called \emph{kinds}, $\kappa$ to distinguish them from  \emph{types}, which are just one kind  of type-level value.
 We will write the kind of types as $\kTypeBlur$, though it is also written $\star$, \verb|T| or \verb|Type| in various similarly structured languages (see Sec.
@@
 In most languages, types are formed by applying one of a fixed collection of \emph{type constructors} to zero or more \emph{indices}.
-In @$\lambda$, the situation is different.
-User-defined type constructors are declared at the top-level of a program (or lifted there, in a practical implementation) using \textsf{tycon} (Fig.
+In @$\lambda$, user-defined type constructors can be declared at the top-level of a program (or lifted there, in a practical implementation) using \textsf{tycon} (Fig.
 \ref{kindprog}).
 Each type constructor in the program must have a unique name, written e.g.
-\fvar{Nat}.
-We assume naming conflicts can be avoided by some extrinsic mechanism (e.g.
-a URI-based scheme).
-A type constructor must also declare an \emph{index kind}, $\kappaidx$.
-Types themselves -- type-level values of kind $\kTypeBlur$ -- are introduced by applying a previously declared type constructor to a type-level term of its index kind, $\ttype{Tycon}{\tauidx}$.
-The kind $\kTypeBlur$ also has an elimination form: a type can be case analyzed against a type constructor in scope to extract its index.
+\fvar{Nat}.\footnote{We assume naming conflicts can be avoided by some extrinsic mechanism.} A type constructor must also declare an \emph{index kind}, $\kappaidx$.
+Introducing a type is thus type constructor application to an index, written $\ttype{Tycon}{\tauidx}$.
+The kind $\kTypeBlur$ also has an elimination form: a type can be case analyzed against an available type constructor to extract its index.
 We are, to a first approximation, treating type constructors as constructors of a built-in open datatype at the type-level.
@@
  To permit the implementation of interesting type systems, the type-level language includes several kinds other than $\kTypeBlur$.
-We lift several  standard functional constructs to the type level: unit ($\kunit$), binary sums ($\ksum{\kappa_1}{\kappa_2}$), binary products ($\kpair{\kappa_1}{\kappa_2}$), lists ($\klist{\kappa}$) and integers ($\dint$).
+We lift several  standard functional fragments to the type level: unit ($\kunit$), binary sums ($\ksum{\kappa_1}{\kappa_2}$), binary products ($\kpair{\kappa_1}{\kappa_2}$), lists ($\klist{\kappa}$) and integers ($\dint$).
 We also include labels ($\klabel$), written in a slanted font, e.g.
-$\tlabel{myLabel}$, which are string-like values that only admit comparisons, here only for equality, and play a distinguished role in the expanded syntax, as we will later discuss.
+$\tlabel{myLabel}$, which are string-like values that only admit comparisons, here only for equality, and play a distinguished role in the expanded syntax, as we will later discuss (strings themselves could also be included, but we omit them for concision).
 Our first example, $\fvar{Nat}$, is indexed trivially, i.e.
@@
 All combinations of kinds introduced thus far  have this property and this is sufficient for many interesting examples.
+Our treatment of equivalence in the type-level language is thus quite similar to the treatment of term-level equality using ``equality types'' in a language like Standard ML.
+An \emph{equality kind} is defined by the judgement \todo{eq type discussion}.
 We discuss introducing a richer notion of  equivalence of types that  fragment providers can extend as future work in Sec.
@@
 The type-level language does, however, include total functions of conventional arrow kind, written $\karrow{\kappa_1}{\kappa_2}$.
-Type formation can be wrapped in a type-level function to emulate a first-class type constructor (indeed, such a wrapper could be generated automatically, though we avoid this for simplicity in our semantics).
-Equivalence at arrow kind does not coincide with $\beta$-equivalence, so type-level functions cannot appear in type indices.
-Our treatment of equivalence in the type-level language is thus quite similar to the treatment of term-level equality using ``equality types'' in a language like Standard ML.
-An equality type is defined by the judgement \todo{eq type discussion}.
-Every type constructor also defines a \emph{representation schema} using \textsf{schema}, a type-level function that we will discuss after introducing operators.
-
-\subsection{Operators}
-For reasons that we will discuss, our calculus associates every operator  constructor\footnote{We distinguish \emph{term constructors} from \emph{type constructors} and \emph{operator constructors}.
-There are term constructors at all levels in the calculus.
-For example, type constructor application is the sole type-forming type-level term constructor.
-An abstract syntax at the type level like the one used by Harper \cite{pfpl} might write it $\texttt{tcapp}[\fvar{Tycon}](\tau)$, emphasizing this point.} with a type constructor.
+Type constructor application can be wrapped in a type-level function to emulate a first-class type constructor (indeed, such a wrapper could be generated automatically, though we avoid this for simplicity in our semantics).
+Equivalence at arrow kind does not meet our criteria, so type-level functions cannot appear within type indices.
+This also prevents general recursion from arising at the type level.
+Without this restriction, a function of type $\karrow{\kTypeBlur}{\kappa}$ could ``smuggle'' in a self-reference as a type index and extract it via case analysis (an issue related to the positivity condition for inductive datatypes in total functional languages).
+
+Every type constructor also defines a \emph{representation schema} using \textsf{schema}, a type-level function that we will discuss in Sec. \ref{?}\todo{ref} after introducing operator constructors.
+
+\subsection{Operator Constructors}
+For reasons that we will discuss, our calculus associates every operator  constructor\footnote{It may be helpful to emphasize that \emph{term formers} are distinct from type constructors and operator constructors.
+There are term formers at all levels in the calculus.
+For example, type constructor application is a type-level term former.
+An abstract syntax at the type level like Harper's \cite{pfpl} might write it $\texttt{tcapp}[\fvar{Tycon}](\tau)$, emphasizing the distinction.} with a type constructor.
 
@@
 For example, the projection operator for labeled tuples, $\fvar{LabeledTuple}.\opvar{prj}$, has index kind $\klabel$.
+\todo{defne operator}
 Note that neither operator constructors nor operators are first-class type-level values and we do not need to impose any restrictions on their index kinds.
@@
 For example, on line 18 of Fig.
-\ref{nat}, we see the operator constructors $\fvar{Nat}.\opvar{z}$ and $\fvar{Nat}.\opvar{s}$ being invoked to compute two (the syntactic overhead will be relieved later).
+\ref{nat}, we see the operator constructors $\fvar{Nat}.\opvar{z}$ and $\fvar{Nat}.\opvar{s}$ being invoked to compute $two$ (the syntactic overhead will be relieved later).
 %\footnote{Although our focus here is entirely on semantics, a brief note on syntax: in the expanded syntax, the trivial indices and empty argument lists can be omitted, so we could write \texttt{Nat.s(Nat.s(Nat.z))}.

### 2014-03-01 att-icfp14/att-icfp14.tex
Exposition finally coming toghetr....

@@
 
+\newcommand{\compresslist}{
+  \vspace{-1em}
+  \setlength{\itemsep}{1pt}
+  \setlength{\parskip}{0pt}
+  \setlength{\parsep}{0pt}
+}
+
 \usepackage{url}
@@
 
-\title{Active Typing: Conservatively Extending a Type System From Within}
+\title{Active Embeddings: Conservatively Extending a Type System From Within}
 
@@
 Type-level functions associated with operator constructors define their static and dynamic semantics, the latter by translation to a fixed typed internal language.
-By lifting compiler correctness techniques pioneered by the TIL compiler for Standard ML into the language, the ``actively typed semantics'' guarantees type safety.
-Going further, the semantics enforce a novel form of representation independence at extension boundaries that ensures that extensions are mutually conservative (i.e.
+By lifting compiler correctness techniques into the language, the ``actively typed semantics'' guarantees type safety.
+Going further, the semantics enforce an abstraction barrier at extension boundaries that ensures that extensions are mutually conservative (i.e.
 they do not weaken or interfere with one another, so that they can  be used together in any combination).
@@
 We intend @$\lambda$ as a minimal foundation for future work on safe language-integrated extension mechanisms for typed programming languages, but it is already quite expressive.
-We demonstrate by showing how a conventional concrete syntax can be introduced by type-directed dispatch to Core @$\lambda$, then discuss a number of typed language fragments that can be introduced as orthogonal ``libraries'' (and discuss the sorts of fragments that, as yet, cannot).
+We demonstrate by showing how a conventional concrete syntax can be introduced by type-directed dispatch to Core @$\lambda$, then discuss a number of typed language fragments that can be \emph{actively embedded} by this mechanism as orthogonal ``libraries''.
+%(and discuss the sorts of fragments that, as yet, cannot).
 \end{abstract}
@@
 The simply typed lambda calculus (STLC), for example, consists of a single fragment containing a single type constructor, $\rightarrow$, indexed by a pair of types, and two operator constructors: $\lambda$, indexed by a type, and \verb|ap|, which we may think of as being indexed trivially.
-A fragment is often identified by the type constructor it is organized around, so the STLC might also be called $\mathcal{L}\{\rightarrow\}$, following the notational convention used by Harper \cite{pfpl} and others.
+A fragment is often identified by the type constructor it is organized around, so the STLC is  called $\mathcal{L}\{\rightarrow\}$, following the notational convention used by Harper \cite{pfpl} and others.
 
@@
 We might thus call it $\mathcal{L}\{\rightarrow \mathtt{nat}\}$.
-This language is  more powerful than the STLC because the STLC admits an embedding (trivially) into  $\mathbf{T}$ but the reverse is not true.
+This language is  more powerful than the STLC because the STLC admits an embedding  into  $\mathbf{T}$ but the reverse is not true.
 Buoyed by this fact, we might go on by adding fragments that define sums, products and various forms of inductive types (e.g.
@@
 If we consider the $\forall$ fragment, however, defining universal quantification over types (i.e.
-parametric polymorphism), we might take pause.
-In $\mathcal{L}\{\rightarrow\,\forall\}$, studied variously by Girard as System $\mathbf{F}$ \cite{girardF} and Reynolds as the polymorphic lambda calculus \cite{polylam}, it is known that sums, products, and inductive and co-inductive types can all be weakly defined via Church-style encodings \cite{reynolds}\todo{what to cite for this?}.
-This fact is of substantial theoretical importance and is quite relevant if \emph{translating} a language like $\mathcal{L}\{\rightarrow\forall~\mathtt{nat} + \times\}$ to $\mathcal{L}\{\rightarrow\forall\}$ is one's main interest.
-But Reynolds, in a remark that recalls the ``Turing tarpit'' of Perlis \cite{Perl82a}, cautioned against using the existence of a weak encoding as an argument for programming with the corresponding embedding directly \cite{Reynolds94anintroduction}:
+parametric polymorphism), we must take a moment to reflect.
+In $\mathcal{L}\{\rightarrow\,\forall\}$, studied variously by Girard as System $\mathbf{F}$ \cite{girardF} and Reynolds as the polymorphic lambda calculus \cite{polylam}, it is known that sums, products, and inductive and co-inductive types can all be weakly encoded \cite{reynolds}\todo{what to cite for this?}.
+This means that we can \emph{translate} well-typed terms of a language like $\mathcal{L}\{\rightarrow\forall~\mathtt{nat} + \times\}$ to well-typed terms in $\mathcal{L}\{\rightarrow\forall\}$ in a manner that preserves their dynamic semantics.
+But Reynolds, in a remark that recalls the ``Turing tarpit'' of Perlis \cite{Perl82a}, reminds us that programming with the corresponding \emph{embedding}, particularly one where only the dynamic semantics are preserved, may be unwise \cite{Reynolds94anintroduction}:
 \begin{quote}
@@
 %Having established why a minimal language like $\mathcal{L}\{\rightarrow\forall\}$, while suitable as an internal language, needs to be extended with additional fragments before it is suitable for use as a human-facing external language,
-Modern ``general-purpose'' languages like ML, on the other hand, expose constructs, like datatypes and abstract types, that admit strong embeddings of many simpler fragments, occupying what is widely seen as a ``sweet spot'' in the design space.
-However, ``general-purpose'' and ''all-purpose'' are quite distinct, and situations continue to arise in both research and practice where introducing a new fragment would be desirable because it can still only be weakly defined in terms of existing fragments, or a strong embedding, while possible, is impractically verbose or inefficient:
+\todo{revise per title}A strong embedding, i.e.
+an isomorphism that preserves static reasoning principles, in a sense that we will make more precise as we go on, is generally far more difficult to establish than a weak embedding.
+For example, the aforementioned embeddings into $\mathcal{L}\{\rightarrow\forall\}$ do not preserve type disequality and other equational principles.
+Harder still is establishing a strong embedding that also supports an acceptably natural programming style (measured, to first approximation, by the amount of boilerplate code needed) and a reasonable cost semantics.
+
+Modern languages like ML do, of course, provide constructs, like datatypes and abstract types, that admit strong and often satisfying embeddings of many language fragments, occupying what is widely seen as a ``sweet spot'' in the design space.
+However, ``general-purpose'' and ''all-purpose'' remain quite distinct, and situations continue to arise in both research and practice where desirable fragments can still only be weakly defined in terms of general-purpose constructs, or a strong embedding, while possible, is widely seen as too verbose or inefficient.
+For example:
+
 \begin{enumerate}
-\item General-purpose abstractions continue to evolve and admit variants.
-There are many  variations on record types, for example: records with functional record update, specified field ordering (that is, labeled tuples) or prototypic delegation are either awkward to strongly embed or cannot be.
-Similarly, sum types also admit variants (e.g.
-various forms of open sums).
-Even something as seemingly simple as a type-safe \verb|sprintf| operator requires extending a language like Ocaml  \cite{ocaml-printf}.
-
+%\compresslist
+\item General-purpose constructs continue to admit variants.
+There are many  variants of product types: records, records with functional record update, labeled tuples (records that specify a field ordering) and records with delegation are either awkward to strongly embed or cannot be.
+Similarly, sum types also admit many variants (e.g.
+various forms of open sums and object systems, seen in a certain light).
+Even something as seemingly simple as a type-safe \verb|sprintf| operator requires special support from a language, e.g.
+Ocaml \cite{ocaml-printf}.
 \item Perhaps more interestingly, specialized type systems that enforce stronger invariants than general-purpose constructs are capable of enforcing are often developed by researchers.
 One need take only a brief  excursion through the literature to discover language extensions that support parallel programming \cite{a}\todo{citations}, concurrent programming \cite{cml}, distributed programming \cite{tom7}, dataflow programming \cite{reactiveml}, authenticated data structures \cite{popl13}, information flow security \cite{walker00, smith2001}, database queries \cite{db}, aliased references \cite{naden12}, network protocols \cite{sekar99}, units of measure \cite{keneddy} and many others.% All of these are implemented as dialects of existing languages, presumably because a strong encoding was not feasible.
-\item Foreign function interfaces (FFIs) that allow programmers to safely and naturally interact with libraries written in an existing language require enforcing the type system of the foreign language within the calling language.
+\item Safe and natural foreign function interfaces (FFIs) require enforcing the type system of the foreign language within the calling language.
 %Using  a FFI that does not do this can lead to safety issues, even when both languages are separately known to be safe.
-Safe FFIs generally require direct extensions to the language.
+%Safe FFIs generally require direct extensions to the language. 
 For example, MLj extended Standard ML with constructs for safely and naturally interfacing with Java \cite{mlj}.
-For any other language, including  others on the JVM like Scala and the many languages that might be accessible via Java's native FFI, there is no way to guarantee that language-specific invariants are statically maintained, and the interface is certainly far from natural.
+For any other language, including  others on the JVM like Scala and the many languages that might be accessible via a native FFI, there is no way to guarantee that language-specific invariants are statically maintained, and the interface is certainly far from natural.
 \end{enumerate}
@@
 This, we argue, is quite unsatisfying: a programmer can choose either a dialect supporting an innovative approach to parallel programming or one that builds in support for statically reasoning about units of measure, but there may not be an available dialect supporting both.
-Forming such a dialect is alarmingly non-trivial, even in the rare situation where a common framework has been used or the dialects are implemented within the same compiler, as these mechanisms do not guarantee that different combinations of individually sound language fragments remain sound when combined.
-Metatheoretic and compiler correctness results can only be derived for the dialect \emph{resulting} from a language   composition operation, so in a diverse fragment ecosystem, fragment providers have little choice but to leave this to clients.
-Avoiding language composition is difficult, as interoperability between components of an application written in different dialects faces precisely the problems described in item 3.% We will consider this further in our discussion of related work in Sec.
+Forming such a dialect is alarmingly non-trivial, even in the rare situation where a common framework has been used or the dialects are implemented within the same compiler, as these mechanisms do not guarantee that different combinations of individually sound dialects remain sound when combined.
+Metatheoretic and compiler correctness results can only be derived for the dialect \emph{resulting} from a language   composition operation, so in a diverse software ecosystem, fragment providers have little choice but to leave this task to clients (providing, perhaps, some informal guidelines or partial automation).
+Avoiding language composition is difficult, as interoperability between components of an application written in different dialects faces precisely the problems of item 3.% We will consider this further in our discussion of related work in Sec.
 \ref{related-work}.
@@
 We draw inspiration from this approach to design a mechanism that makes it possible to declare new type and operator constructors and define, \emph{using type-level functions}, their static and dynamic semantics, the latter by translation to a fixed typed internal language.
-If a language fragment can be weakly defined in terms of the typed internal language and its statics are of a form that can be implemented by this protocol, it can be made strongly embeddable in the external language by introducing, in this way, the necessary constructors and static logic.
-
-The semantics imposes checks on extensions to guarantee type safety and decidability of typing and type equality.
-Moreover, an important class of lemmas that play a key role in  proofs that an extension enables strong embeddings of a language fragment can be derived assuming a ``closed world'' (that is, in the completely traditional manner where one constructs a complete language and reasons inductively).
-The semantics guarantees that they will be \emph{conserved} in the  ``open world'' by enforcing a form of  representation independence at all extension boundaries.
-This avoids the biggest problems of composing language dialects and justifies the inclusion of the mechanism directly into the language.
+If a language fragment can be weakly defined in terms of the typed internal language and its statics are of a form that can be implemented by this protocol, it can be made strongly embeddable in the external language by introducing, in this way, the necessary constructors and type assignment logic.
+We call such an embedding an \emph{active embedding}.
+
+The semantics imposes checks on extensions to guarantee type safety and decidability of type assignment and equality.
+Moreover, an important class of lemmas that play a key role in  proofs that an extension enables strong embeddings of a language fragment can be derived in a suitable ``closed world'' (that is, in the traditional manner where one reasons inductively over a language composed of a minimal collection of fragments).
+The semantics guarantees that they will be \emph{conserved} in the  ``open world'' by enforcing a form of  representation independence at extension boundaries.
+This avoids the most fundamental problems of composing dialects and justifies the inclusion of the mechanism inside the language.
 
 We will begin in Sec.
-\ref{core} by introducing a minimal calculus with this kind of \emph{actively typed semantics}, @$\lambda$.
-The external language begins with only the $\rightarrow$ fragment and the typed internal language is a simple variant of PCF that permits only weak embeddings of even fragments like the $\mathtt{nat}$ fragment of G\"odel's $\mathbf{T}$ or ML-style $n$-ary products.
+\ref{core} by introducing a minimal calculus with this kind of {actively typed} semantics, @$\lambda$.
+The external language begins with only the $\rightarrow$ fragment and the typed internal language is a simple variant of PCF that permits only weak embeddings of even simple fragments like the $\mathtt{nat}$ fragment of G\"odel's $\mathbf{T}$ or ML-style $n$-ary products.
 We implement these as extensions to explain the mechanism and motivate the constraints that the semantics imposes.
@@
 \ref{theory}.
-These are the fundamental contributions of this work.
-
-While the core calculus addresses the issue of making a strong embedding possible, embeddings into the core calculus are quite clearly impractical syntactically.
+These two sections are the fundamental contributions of this work.
+
+While the core calculus addresses the issue of making a strong embedding possible, the embedding into the core calculus is quite clearly impractical syntactically.
 To begin to address the issue of syntax, we develop in Sec.
-\ref{expanded-syntax} a flexible type-directed dispatch protocol for a more conventional \emph{expanded syntax} that defers to the core calculus semantically.
+\ref{expanded-syntax} a flexible type-directed dispatch protocol for a more conventional {expanded syntax} that defers to the core calculus semantically.
 We then use the expanded syntax to discuss several more interesting examples in Sec.
-\ref{examples}: \todo{list these}.
-\todo{organization of rest of sections}.
-
-While of surprising expressiveness, @$\lambda$ itself is certainly not capable of expressing the full variety of type systems listed earlier.
+\ref{examples}: labeled sums and products, the latter with delegation and functional field update, and a variant of \verb|sprintf|.\todo{update this if something doesn't make it}
+
+While of surprising expressiveness given its minimality, @$\lambda$ itself is certainly not capable of admitting embeddings of the full variety of type system fragments enumerated earlier.
 Indeed, to guarantee safety and conservativity, our calculus assumes perhaps too much about the type system being encoded.
-For example, it only admits fragments for which the typing judgement looks essentially like the typing judgement of the $\rightarrow$ fragment; new contexts cannot be introduced.
-We discuss this and some of the other reasons why a fragment may not be expressible in @$\lambda$ in Sec.
+For example, it only admits fragments where the typing judgement looks essentially like the typing judgement of the $\rightarrow$ fragment, so new contexts cannot be introduced.
+We discuss this and some of the other reasons why a fragment may not be satisfyingly embeddable in Sec.
 \ref{limitations}.
-Fortunately, these limitations do not appear fundamental to the approach, so we also outline in Sec.
+These limitations do not appear fundamental to the approach, so we also outline in Sec.
 \ref{limitations} directions for future work.
-We conclude with a discussion of related work in Sec.
+We conclude with comparisons to related work in Sec.
 \ref{related-work}.
@@
 \ref{grammar}.
-An example of a program defining type and operator constructors that can be used to strongly embed G\"odel's \textbf{T} in @$\lambda$ is given in Fig.
+An example of a program defining type and operator constructors that can be used to strongly encode G\"odel's \textbf{T} in @$\lambda$ is given in Fig.
 \ref{nat}.
-We will discuss its semantics and how precisely the strong embedding, seen being used starting on line 16, works as we go on.
+We will discuss its semantics and how precisely the corresponding embedding, seen being used starting on line 16, works as we go on.
 Natural numbers can, of course, be strongly embedded in existing languages with a similar usage and asymptotic performance profile (up to function call overhead with an abstract type, for example).
-We will provide more sophisticated examples where this is less feasible later on.
-%It consists of:
+We will provide more sophisticated examples where this is less feasible later on (and discuss type abstraction itself in Sec.
+\ref{limitations}).\todo{if i don't have time/space, take this out} %It consists of:
 %a type constructor declaration, $\fvar{Nat}$, indexed trivially, together with three operator constructors, also all indexed trivially, that implement the standard introductory forms for natural numbers as well as the recursor operator (as in G\"odel's T \cite{pfpl}).
@@
 
-\subsection{Programs}
+\subsection{Overview}\label{programs}
 \begin{figure}[t]
@@
 \caption{\small Syntax of Core \atlam.
-Here, $x$ ranges over external and internal language variables, $\tvar{t}$ ranges over type-level variables, $\fvar{Tycon}$ ranges over type constructor names, $\opvar{op}$ ranges over operator constructor names, $\iintlit$ ranges over integer literals, $\tlabel{label}$ ranges over label literals (see text) and $\oplus$ ranges over standard total binary operations over integers (e.g.
-addition).%The productions related to the internal language are written using generators $\mathcal{G}$ and $\mathcal{S}$ to avoid duplicating the syntax of common terms.
-\vspace{-10pt}
+Here, $x$ ranges over external and internal language variables, $\tvar{t}$ ranges over type-level variables, $\fvar{Tycon}$ ranges over type constructor names, $\opvar{op}$ ranges over operator constructor names, $\iintlit$ ranges over integer literals, $\tlabel{label}$ ranges over label literals (see text) and $\oplus$ ranges over standard total binary operations (e.g.
+addition, comparison).
 \label{grammar}}
@@
 \\
-&}{XXX}{i}{\tlam{idx}{\kunit}{\titype{\dint}}};\\
+&}{XXX}{i}{\tlam{i}{\kunit}{\titype{\dint}}};\\
 \label{nattype}&{\sf let}~\tvar{nat} = \ttype{Nat}{\tunit}~{\sf in}\\
@@
 %\vspace{-15px}
-\caption{\small G\"odel's $\mathbf{T}$ in \atlam, used to calculate 2+2.
-The simple helper functions $\tvar{arity}n$ and $\tvar{ifeq}$ have return kind $\ksum{\kDen}{\kunit}$ (see text).
-Their definitions can be found in the appendix.
-We use \textsf{let} to bind both external and type-level variables for clarity of presentation; these can be eliminated manually.
-We will discuss the issue of a more natural syntax in Sec.
+\caption{\small An embedding of G\"odel's $\mathbf{T}$ in \atlam, used to define $plus$ and $two$ and calculate $plus~two~two$.
+The simple helper functions $\tvar{arity}n$ and $\tvar{ifeq}$ have return kind $\ksum{\kDen}{\kunit}$ and should be replaced by the definitions given in Fig.
+\ref{helpers}.
+We use \textsf{let} to bind both external and type-level variables for clarity of presentation.
+We will discuss further syntactic issues in Sec.
 \ref{expanded-syntax}.}
 \label{nat}
+\vspace{-10pt}
 \end{figure}
 A \emph{program}, $\rho$, consists of a series of static declarations (in the core, only constructor declarations) followed by an external term, $e$.
-The syntax for external terms contains three term formers: variables, $\lambda$, and a form for all other operator invocations, which we will discuss below.
-Compiling a program consists of first \emph{kind checking} its static declarations and type-level terms (Fig.
-\ref{kindprog}, discussed below) then \emph{type checking} the external term and, simultaneously, \emph{translating} it to a term, $\iota$, in the {typed internal language} (Fig.
+The syntax for external terms contains three forms: variables, $\lambda$ terms, and a form for invoking user-defined operators, which we will discuss below.
+Compiling a program consists of first \emph{kind checking} its static declarations and type-level terms, $\tau$, which we will describe in Sec \ref{types} (Fig.
+\ref{kindprog}), then typechecking the external term and simultaneously {translating} it to a term, $\iota$, in the {typed internal language}, described thereafter (Fig.
 \ref{att}).
-The key judgement is the \emph{active typing judgement}, relating an external term to a {type} and an internal term, called its \emph{translation}, under \emph{typing context} $\Gamma$ and \emph{constructor context} $\Phi$:
-$$\ecompilesX{e}{\tau}{\iota}$$
-
-This judgement follows standard conventions for the typing judgement of a simply-typed lambda calculus in most respects.
-The typing context $\Gamma$ maps variables to types and obeys standard structural properties.
-The key novelties here are 1) that the available type and operator constructors are not fixed by the language, but rather are tracked by the constructor context, $\fvalCtx$, and 2) that a translation is simultaneously derived.
-There is no separate operational semantics for the external language; the dynamic semantics are defined by this translation to the internal language (which might have a more conventional operational semantics).
-\todo{cite Harper-Stone / denotational semantics} The two contexts and the external term can be thought of as input to the judgement, while the type and translation are output (we do not consider a bidirectional approach \cite{LovasPfenning}\todo{cite} or type inference in this work, though this is likely a promising avenue for future research).
-
-In our calculus, the internal language (IL), consisting of internal terms $\iota$ and internal types $\sigma$, provides partial functions (via the generic fixpoint operator of Plotkin's PCF \cite{PCF}), simple product and sum types and a base type of integers (to make our example interesting and as a nod toward speed on contemporary machines).
-In practice, the internal language could be any typed  language with a specification for which type safety and decidability of typechecking have been satisfyingly determined (e.g.
-Standard ML or CakeML \cite{cakeml}, leaving translation to a simpler IL to later stages of the compilation process).
-
-\subsection{Types and Type Constructors}
+We can thus specify the \emph{central compilation judgement} $\pkcompiles{\rho}{\iota}$ with one rule:
+\[
+\inferrule[p-compiles]{
+	\progOK{\emptyset}{\fvalCtx_0}{\rho}\\
+	\pcompiles{\fvalCtx_0}{\rho}{\iota}
+}{\pkcompiles{\rho}{\iota}}
+\]
+The key judgement form in the calculus is the \emph{active typing judgement}  (Fig.
+\ref{att}, which we will describe in Secs.
+\ref{opcons}-\ref{repind}), relating an external term to a {type}, $\tau$, called its \emph{type assignment}, and an internal term, called its \emph{translation}, under \emph{typing context} $\Gamma$ and \emph{constructor context} $\Phi$:
+\[\ecompilesX{e}{\tau}{\iota}\]
+
+The typing context $\Gamma$ is essentially conventional, mapping variables to types and obeying standard structural rules (weakening, exchange and contraction; \cite{pfpl} describes the necessary background for this paper).
+The constructor context tracks user-defined type and operator constructors, the topics of the remainder of this section, and establishing its structural properties requires establishing the conservativity properties that we will discuss in Sec.
+\ref{theory}.% Weakening and exchange of the constructor context closely related to the issue of conservativity that we will return to.
+Each constructor is identified by name, so there is no analog to contraction.
+
+There is no separate operational semantics for the external language; the dynamic behavior of an external term is determined by its translation to the internal language (which might have a more conventional operational semantics).
+This form of semantics can be considered as a lifting into the language specification of the first stage of a type-directed compiler like the TIL compiler for Standard ML \cite{TIL} and has some parallels to the Harper-Stone semantics for Standard ML, where external terms were also given meaning by elaboration to an IL \cite{harper-stone}\todo{citations}.
+%
+
+In @$\lambda$, the internal language (IL) provides partial functions (via the generic fixpoint operator of Plotkin's PCF), simple product and sum types and a base type of integers (to make our example interesting and as a nod toward speed on contemporary machines).
+In practice, the internal language could be any typed  language with a specification for which type safety and decidability of typechecking have been satisfyingly determined.
+Constraints that must be maintained for all well-typed terms, independent of the constructor context (e.g.
+guarantees that out-of-bounds access to memory never occurs), must be maintained by the internal type system and program performance is ultimately limited by the internal language and downstream compilation stages that we do not here consider (safely extending downstream compilation has been discussed in previous work, e.g.
+\cite{compexts}).
+
+\subsection{Type Constructors and Types (Fig. \ref{nat}, lines 1 and 16)}\label{types}
 \begin{figure}[t]
@@
 $\fbox{\inferrule{}{\progOKX{\progsort}}}$
-%~~~$\tvarCtx ::= \emptyctx \pipe \tvarCtxX{t}{\kappa}$~~~
+~~~$\fvalCtx ::= \emptyctx \pipe \fvalCtxX{\fvalDf}$
+~~~$\tvarCtx ::= \emptyctx \pipe \tvarCtxX{t}{\kappa}$
 %$\fCtx ::= \Sigma_0	 \pipe \fvalCtxX$
@@
 \end{mathpar}
-\caption{\small Kinding for programs.
-The kinding context $\tvarCtx$ maps type variables to kinds (we do not specify let binding of type variables at the level of programs or external terms in the core, but this would be a simple addition).
-Type-level terms stored in the constructor context are not needed during kinding, indicated using a dash.
-Kinding  for type-level terms is specified in Figure \ref{tlkind}.}
+$\fbox{$\kEq{\kappa}$}$
+\begin{mathpar}
+\inferrule[t-eq]{ }{
+	\kEq{\kTypeBlur}
+}
+
+\inferrule[i-eq]{ }{
+	\kEq{\kint}
+}
+
+\inferrule[l-eq]{ }{
+	\kEq{\klabel}
+}
+
+\inferrule[list-eq]{
+	\kEq{\kappa}
+}{
+	\kEq{\klist{\kappa}}
+}
+\\
+\inferrule[u-eq]{ }{
+	\kEq{\kunit}
+}
+
+\inferrule[p-eq]{
+	\kEq{\kappa_1}\\
+	\kEq{\kappa_2}
+}{
+	\kEq{\kpair{\kappa_1}{\kappa_2}}
+}
+
+\inferrule[s-eq]{
+	\kEq{\kappa_1}\\
+	\kEq{\kappa_2}
+}{
+	\kEq{\ksum{\kappa_1}{\kappa_2}}
+}
+\end{mathpar}
+\caption{\small Kinding for programs, operator constructors and external terms.
+The kinding context $\tvarCtx$ maps type variables to kinds (we do not specify let binding of type-level variables at any level in the core, but this would be a simple addition).
+Type-level terms stored in the constructor context are not needed during kinding, indicated using a dash.}
 \label{kindprog}
@@
 \small
+%\vspace{-15pt}
+$\fbox{\inferrule{}{\tKindX{\tau}{\kappa}}}$
+\begin{mathpar}
+\small\inferrule[k-var]{
+}{
+  \tKind{\tvarCtxX{t}{\kappa}}{\fvalCtx}{\tvar{t}}{\kappa}
+}
+~~~~
+\inferrule[k-arrow-i]{
+  \tKind{\tvarCtxX{t}{\kappa_1}}{\fvalCtx}{\tau}{\kappa_2}
+}{
+  \tKindX{\tlam{t}{\kappa_1}{\tau}}{\karrow{\kappa_1}{\kappa_2}}
+}
+~~~~
+\inferrule[k-arrow-e]{
+  \tKindX{\tau_1}{\karrow{\kappa_1}{\kappa_2}}\\\\
+  \tKindX{\tau_2}{\kappa_1}
+}{
+  \tKindX{\tapp{\tau_1}{\tau_2}}{\kappa_2}
+}
+
+\text{\color{gray} (kinding for integers, labels, lists, products and sums also standard)}
+
+%%
+%%\inferrule{ }{
+%%	\tKindX{\tstr{str}}{\kstr}
+%%}(\text{str}^I_\tau)
+%%
+%%\inferrule{ }{
+%%	\tKindX{\tunit}{\kunit}
+%%}(\text{1}^I_\tau)
+%%
+%%\inferrule{
+%%	\tKindX{\tau_1}{\kappa_1}\\
+%%	\tKindX{\tau_2}{\kappa_2}
+%%}{
+%%	\tKindX{\tpair{\tau_1}{\tau_2}}{\kpair{\kappa_1}{\kappa_2}}
+%%}({\times}^I_\tau)
+%%
+%%\inferrule{
+%%	\tKindX{\tau}{\kpair}{\kappa_1}{\kappa_2}
+%%}{
+%%	\tKindX{\tfst{\tau}}{\kappa_1}
+%%}({\times}^{E1}_\tau)
+%%
+%%\inferrule{
+%%	\tKindX{\tau}{\kpair}{\kappa_1}{\kappa_2}
+%%}{
+%%	\tKindX{\tsnd{\tau}}{\kappa_2}
+%%}({\times}^{E2}_\tau)
+%%\inferrule{ }{
+%%\inferrule{
+%%	\tKindX{\tau_{1}}{\kappa_{1}}\\
+%%	\tKindX{\tau_{2}}{\kappa_{2}}
+%%}{
+%%	\tKindX{\tpair{\tau_{1}}{\tau_{2}}}{\kpair{\kappa_{1}}{\kappa_{2}}}
+%%}~(\times_\tau)
+%%
+%%\inferrule{
+%%	\tKindX{\tau}{\kpair{\kappa_{1}}{\kappa_{2}}}
+%%}{
+%%	\tKindX{\tfst{\tau}}{\kappa_{1}}
+%%}~(\text{fst}_\tau)
+%%
+%%\inferrule{
+%%	\tKindX{\tau}{\kpair{\kappa_{1}}{\kappa_{2}}}
+%%}{
+%%	\tKindX{\tsnd{\tau}}{\kappa_{2}}
+%%}~(\text{snd}_\tau)
+%%
+%% TODO: list
+\inferrule[k-eq]{
+	\kEq{\kappa}\\
+	\tKindX{\tau_1}{\kappa}\\
+	\tKindX{\tau_2}{\kappa}\\\\
+	\tKindX{\tau_3}{\kappa'}\\
+	\tKindX{\tau_4}{\kappa'}
+}{
+	\tKindX{\tifeq{\tau_1}{\tau_2}{\kappa}{\tau_3}{\tau_4}}{\kappa'}
+}
+
+\inferrule[k-ty-i]{
+	\fval{Tycon}{\kappaidx}{-}{-} \in \fvalCtx\\\\
+	\tKind{\tvarCtx}{\fvalCtx}{\tauidx}{\kappaidx}
+}{
+	\tKindX{\ttype{Tycon}{\tauidx}}{\kTypeBlur}
+}
+
+\inferrule[k-ty-e]{
+	\tKindX{\tau}{\kTypeBlur}\\
+	\fval{Tycon}{\kappaidx}{-}{-} \in \fvalCtx\\\\
+	\tKind{\tvarCtxX{x}{\kappaidx}}{\fvalCtx}{\tau_1}{\kappa}\\
+	\tKindX{\tau_2}{\kappa}
+}{
+	\tKind{\tvarCtx}{\fvalCtx}{\tfamcase{\tau}{Tycon}{x}{\tau_1}{\tau_2}}{\kappa}
+}
+~~~~~~
+\inferrule[k-d-i]{
+	\tKindX{\tau}{\kTypeBlur}\\\\
+	\isItermX{\ibar}
+}{
+	\tKindX{\tden{\ibar}{\tau}}{\kDen}
+}
+
+\inferrule[k-d-i-checked]{
+	\tKindX{\tau}{\kTypeBlur}\\\\
+	\isItermX{\ibar}
+}{
+	\tKindX{\tden{\ibar}{\tau}^\checkmark}{\kDen}
+}
+
+\inferrule[k-d-e]{
+	\tKindX{\tau}{\kDen}
+}{
+	\tKindX{\ttypeof{\tau}}{\kDen}
+}
+%
+%\inferrule[iterm-intro]{
+%	\isIterm{\tvarCtx}{\emptyctx}{\fvalCtx}{\iota}
+%}{
+%	\tKindX{\titerm{\iota}}{\kITerm}
+%}
+
+\inferrule[k-reptype-i]{
+	\isItype{\tvarCtx}{\fvalCtx}{\sbar}
+}{
+	\tKindX{\titype{\sbar}}{\kIType}
+}
+\end{mathpar}
+$\fbox{$\isItermX{\ibar}$}$
+\begin{mathpar}
+\inferrule[k-i-var]{ }{
+	\isItermX{\evar{x}}
+}
+
+\inferrule[k-i-fix]{
+	\isItypeX{\sbar}\\\\
+	\isItermX{\ibar}
+}{
+	\isItermX{\ifix{\evar{x}}{\sbar}{\ibar}}
+}
+
+\inferrule[k-i-transof]{
+	\tKindX{\tau}{\kDen}
+}{
+	\isItermX{\itransof{\tau}}
+}
+%\inferrule[i-lam-kinding]{
+%	\isItypeX{\sbar}\\\\
+%	\isItermX{\ibar}
+%}{
+%	\isItermX{\ilam{\evar{x}}{\sbar}{\ibar}}
+%}
+
+\text{\color{gray} (omitted forms are treated recursively, as exemplified by \textsc{k-i-fix})}
+%\inferrule[iterm-dereify-kinding]{
+%	\tKindX{\tau}{\kITerm}
+%}{
+%	\isItermX{\iup{\tau}}
+%}
+%
+\end{mathpar}
+$\fbox{$\isItypeX{\sbar}$}$
+\begin{mathpar}
+%\inferrule[i-int-kinding]{ }{
+%	\isItypeX{\dint}
+%}
+%
+%\inferrule[i-prod-kinding]{
+%	\isItypeX{\sigma_1}\\\\
+%	\isItypeX{\sigma_2}
+%}{
+%	\isItypeX{\dpair{\sigma_1}{\sigma_2}}
+%}
+%
+\inferrule[k-s-parr]{
+	\isItypeX{\sigma_1}\\\\
+	\isItypeX{\sigma_2}
+}{
+	\isItypeX{\darrow{\sigma_1}{\sigma_2}}
+}
+~~~~~~
+\inferrule[k-s-unquote]{
+	\tKindX{\tau}{\kIType}
+}{
+	\isItypeX{\dup{\tau}}
+}
+~~~~~~
+\inferrule[k-s-repof]{
+	\tKindX{\tau}{\kTypeBlur}
+}{
+	\isItypeX{\trepof{\tau}}
+}
+
+\text{\color{gray} (omitted forms are treated recursively, as exemplified by \textsc{k-i-parr})}
+\end{mathpar}
+\caption{\small Kinding for type-level terms. Checked derivates should only be constructed by the semantics (a restriction that can be enforced by a parser).}
+\label{tlkind}
+%\vspace{-15px}
+\end{figure} 
+In most languages, types are formed by applying one of a fixed collection of \emph{type constructors} to zero or more \emph{indices}.
+In @$\lambda$, the situation is notionally similar.
+User-defined type constructors can be declared at the top of a program (or lifted to the top, in practice) using \textsf{tycon}.
+Each constructor in the program must have a unique name, written e.g.
+\fvar{Nat}.\footnote{We assume naming conflicts can be avoided by some extrinsic mechanism.} A type constructor must also declare an \emph{index kind}, $\kappaidx$.
+A type is introduced by applying a type constructor to an index of this kind, written $\ttype{Tycon}{\tauidx}$.
+
+@$\lambda$ supports, and makes extensive use of, simply-kinded type-level computation.
+Specifically, type-level terms, $\tau$, themselves form a typed lambda calculus.
+The classifiers of type-level terms are, following convention, called \emph{kinds}, $\kappa$, to distinguish them from  \emph{types}.
+Types are  type-level values of kind $\kTypeBlur$, introduced as just described.
+The kind $\kTypeBlur$ also has an elimination form, $\tfamcase{\tau}{Tycon}{x}{\tau_1}{\tau_2}$, allowing the extraction of a type index by case analysis against an available type constructor.
+To a first approximation, we are treating type constructors as constructors of a built-in open datatype at the type-level.
+Like open datatypes, there is no notion of exhaustiveness and the default case is required for totality.
+%We will write the kind of types as $\kTypeBlur$, though it is also written $\star$ or \verb|Type| in various similarly structured languages (see Sec. \ref{related-work}). 
+The kinding rules are specified in Fig.
+\ref{kindprog} and Fig.
+\ref{tlkind} and the reader is encouraged to refer to these as we continue.% Rather than there being a fixed set of type constructors, we allow the programmer to declare new type  constructors, and give the static and dynamic semantics of their associated operators, by writing type-level functions.
+In the semantics for this calculus, our kind system combined with techniques borrowed from the typed compilation literature and a form of type abstraction allow us to prove strong type safety, decidability and conservativity theorems.
+
+ To permit the implementation of interesting type systems, the type-level language includes several kinds other than $\kTypeBlur$.
+We lift several  standard functional fragments to the type level: unit ($\kunit$), binary products ($\kpair{\kappa_1}{\kappa_2}$), binary sums ($\ksum{\kappa_1}{\kappa_2}$), lists ($\klist{\kappa}$) and integers ($\dint$).
+We also include labels ($\klabel$), written in a slanted font, e.g.
+$\tlabel{myLabel}$, which are string-like values that only support comparison and play a distinguished role in the expanded syntax, as we will later discuss.
+Our first example, $\fvar{Nat}$, is indexed trivially, i.e.
+by unit kind, $\kunit$, so there is only one natural number type, $\ttype{Nat}{\tunit}$, but we will show examples of type constructors that are indexed in more interesting ways in later portions of this work.
+For example, $\fvar{Tuple}$ has index kind $\klist{\kTypeBlur}$ and $\fvar{LabeledTuple}$ has index kind $\klist{\kpair{\klabel}{\kTypeBlur}}$.
+The type constructor $\fvar{Arrow}$ is included in the initial constructor context, $\fvalCtx_0$, shown in Fig.
+\ref{arrow}, and has index kind $\kpair{\kTypeBlur}{\kTypeBlur}$.
+We will discuss it further below.
+
+Two type-level terms of kind $\kTypeBlur$ are equivalent if they apply the same constructor, identified by name, to equivalent indices.
+Going further, we ensure that deciding type equivalence requires only checking for syntactic equality after normalization by imposing the restriction that equivalence at a type constructor's index kind must be decidable in this way.
+All combinations of kinds introduced thus far  have this property and this is sufficient for many interesting examples.
+Our treatment of equivalence in the type-level language is thus quite similar to the treatment of term-level equality using ``equality types'' in a language like Standard ML.
+A kind $\kappa$ is an  \emph{equality kind} if $\kEq{\kappa}$ can be derived (Fig.
+\ref{kindprog}).
+Conditional branching on the basis of equality at an equality kind can be performed in the type-level language (e.g.
+in \tvar{ifeq}, Fig.
+\ref{helpers}, discussed below).% as maintaining the metatheoretic guarantee that typing respects type equivalence would impose a substantial burden in such a setting.% (a na\"ive approach to this would impose non-trivial extrinsic proof obligations onto extension developers that, unlike in others in this thesis, could threaten type safety).
+
+Type constructors are not first-class; they do not themselves have arrow kind as in some kind systems (e.g.
+\cite{Watkins08}\todo{cite}; Ch.
+22 of \emph{PFPL} describes a related system \cite{pfpl}).
+The type-level language does, however, include total functions of conventional arrow kind, written $\karrow{\kappa_1}{\kappa_2}$.
+Type constructor application can be wrapped in a type-level function to emulate a first-class type constructor (indeed, such a wrapper could be generated automatically, though we avoid this for simplicity in our semantics).
+Equivalence at arrow kind does not meet our criteria, so type-level functions cannot appear within type indices.
+This also prevents general recursion from arising at the type level.
+Without this restriction, a type-level function taking a type as an argument could ``smuggle in'' a self reference as a type index and extract it via case analysis (continuing our analogy to open datatypes, this is closely related to the positivity condition for inductive datatypes in total languages).
+
+Every type constructor also defines a \emph{schema}, a type-level function that associates with every index an internal type.
+The \emph{representation type} of a type like $\ttype{Nat}{\tunit}$ is then determined by applying its constructor's schema to its index, here resulting in $\dint$.
+We will return to this in Sec.
+\ref{repcon} after introducing operators.
+\subsection{Operators and Operator Constructors}\label{opcons}
+\begin{figure}[t]
+\small
+\[
+\begin{array}{rcl}
+\tvar{tyerr} & := & \tinr{\kDen}{\tunit}\\
+\tvar{tyasgn} & := & (\tlam{d}{\kDen}{\tinl{\kunit}{\tvar{d}}})\\
+\tvar{ifeq} & := & (\tlam{t1}{\kTypeBlur}{\tlam{t2}{\kTypeBlur}{\tlam{d}{\kDen}{
+	\\& & \quad \tifeq{\tvar{t1}}{\tvar{t2}}{\kTypeBlur}{\tvar{tyasgn}~\tvar{d}}{\tvar{tyerr}}\\
+\tvar{arity0} & := & (\tlam{a}{\klist{\kDen}}{\tlam{d}{(\ksum{\kDen}{\kunit})}{\tfold{\tvar{a}}{\tvar{d}}{\_}{\_}{\_}{\tvar{tyerr}}}})\\
+\tvar{arity1} & := & (\tlam{a}{\klist{\kDen}}{\tlam{k}{\karrow{\kDen}{(\ksum{\kDen}{\kunit})}}{
+	\\ & & \quad \tfold{\tvar{a}}{\tvar{tyerr}}{h}{t}{\_}{
+ \tvar{arity0}~\tvar{t}~(\tvar{k}~\tvar{h})}}})\\
+\tvar{arity2} & := & (\tlam{a}{\klist{\kDen}}{\tlam{k}{\karrow{\kDen}{\karrow{\kDen}{(\ksum{\kDen}{\kunit})}}}{
+	\\ & & \quad \tfold{\tvar{a}}{\tvar{tyerr}}{h}{t}{\_}{
+ \tvar{arity1}~\tvar{t}~(\tvar{k}~\tvar{h})}}})\\
+ \tvar{arity3} & := & (\tlam{a}{\klist{\kDen}}{\tlam{k}{\karrow{\kDen}{\karrow{\kDen}{\karrow{\kDen}{(\ksum{\kDen}{\kunit})}}}}{
+	\\ & & \quad \tfold{\tvar{a}}{\tvar{tyerr}}{h}{t}{\_}{
+ \tvar{arity2}~\tvar{t}~(\tvar{k}~\tvar{h})}}})
+}}}
+\end{array}
+\]
+\caption{\small Useful (type-level) functions for writing operator definitions.}
+\vspace{-10pt}
+\label{helpers}
+\end{figure}
+\begin{figure}[t]
+\small
+\[
+\begin{array}{rcl}
+\Phi_0 & := & \emptyset, \fval{Arrow}{\kpair{\kTypeBlur}{\kTypeBlur}}{\tau_0}{\theta_0}\\
+\tau_0 & := & \tlam{i}{\kpair{\kTypeBlur}{\kTypeBlur}}{\titype{\darrow{\trepof{\tfst{\tvar{i}}}}{\trepof{\tsnd{\tvar{i}}}}}}\\
+\theta_0 & := & \tops{ap}{\kunit}{i}{a}{
+	\tlam{i}{\kunit}{\tlam{a}{\klist{\kDen}}{
+		\tvar{arity2}~\tvar{a}~\tlam{d1}{\kDen}{\tlam{d2}{\kDen}{
+		\\& & \quad \tfamcase{\ttypeof{\tvar{d1}}}{Arrow}{x}{
+		\\& & \quad\quad \tvar{ifeq}~\tfst{\tvar{x}}~\ttypeof{\tvar{d2}}
+		\\& & \quad\quad\quad \tden{\itransof{\tvar{d1}}~\itransof{\tvar{d2}}}{\tsnd{\tvar{x}}}
+		\\& & \quad\hspace{-3px}}{\tvar{tyerr}}
+		}}
+	}}
+}
+\end{array}
+\]
+\caption{\small The initial constructor context, $\fvalCtx_0$, defines the $\fvar{Arrow}$ type constructor.
+The introductory form, $\elam{x}{\tau}{e}$, is built into the language but the elimination form goes through the operator constructor $\fvar{Arrow}.\opvar{op}$.}
+\label{arrow}
+\end{figure}
+\begin{figure}[t]
+\small
 $\fbox{\inferrule{}{\pcompilesX{\rho}}}$
-~~~$\fvalCtx ::= \emptyctx \pipe \fvalCtxX{\fvalDf}$
 \begin{mathpar}
 \inferrule[att-tycon]{
-	\pcompiles{\fvalCtxX{\fvalDf}}{\rho}{\gamma}
+	\pcompiles{\fvalCtxX{\fvalDf}}{\rho}{\iota}
 }{
@@
 	\ecompiles{\emptyctx}{\fvalCtx}{e}{\tau}{\iota}%\\
-%	\eraseX{\gabs}{\gamma}
+%	\eraseX{\gabs}{\iota}
 }{
@@
 \end{figure}
-@$\lambda$ is an ``actively typed'' version of the simply-typed lambda calculus with simply-kinded type-level computation.
-Specifically, type-level terms, $\tau$, themselves form a typed lambda calculus.
-The classifiers of terms at the type level are called \emph{kinds}, $\kappa$ to distinguish them from  \emph{types}, which are just one kind  of type-level value.
-We will write the kind of types as $\kTypeBlur$, though it is also written $\star$, \verb|T| or \verb|Type| in various similarly structured languages (see Sec.
-\ref{related-work}).
-The kinding rules for type-level terms are given in Fig.
-\ref{todo}\todo{refer}.% Rather than there being a fixed set of type constructors, we allow the programmer to declare new type  constructors, and give the static and dynamic semantics of their associated operators, by writing type-level functions.
-In the semantics for this calculus, our kind system combined with techniques borrowed from the typed compilation literature and a form of type abstraction allow us to prove strong type safety, decidability and conservativity theorems.
-
-In most languages, types are formed by applying one of a fixed collection of \emph{type constructors} to zero or more \emph{indices}.
-In @$\lambda$, user-defined type constructors can be declared at the top-level of a program (or lifted there, in a practical implementation) using \textsf{tycon} (Fig.
+User-defined operator constructors are declared using \textsf{opcon} (cf.
+Fig.
 \ref{kindprog}).
-Each type constructor in the program must have a unique name, written e.g.
-\fvar{Nat}.\footnote{We assume naming conflicts can be avoided by some extrinsic mechanism.} A type constructor must also declare an \emph{index kind}, $\kappaidx$.
-Introducing a type is thus type constructor application to an index, written $\ttype{Tycon}{\tauidx}$.
-The kind $\kTypeBlur$ also has an elimination form: a type can be case analyzed against an available type constructor to extract its index.
-We are, to a first approximation, treating type constructors as constructors of a built-in open datatype at the type-level.
-Like open datatypes, there is no longer a notion of exhaustiveness and the default case is required for case analysis.
-
- To permit the implementation of interesting type systems, the type-level language includes several kinds other than $\kTypeBlur$.
-We lift several  standard functional fragments to the type level: unit ($\kunit$), binary sums ($\ksum{\kappa_1}{\kappa_2}$), binary products ($\kpair{\kappa_1}{\kappa_2}$), lists ($\klist{\kappa}$) and integers ($\dint$).
-We also include labels ($\klabel$), written in a slanted font, e.g.
-$\tlabel{myLabel}$, which are string-like values that only admit comparisons, here only for equality, and play a distinguished role in the expanded syntax, as we will later discuss (strings themselves could also be included, but we omit them for concision).
-Our first example, $\fvar{Nat}$, is indexed trivially, i.e.
-by unit kind, $\kunit$, because there is only one natural number type, $\ttype{Nat}{\tunit}$, but we will show examples of type constructors that are indexed in more interesting ways in later portions of this work.
-For example, $\fvar{Tuple}$ has index kind $\klist{\kTypeBlur}$ and $\fvar{LabeledTuple}$ has index kind $\klist{\kpair{\klabel}{\kTypeBlur}}$.
-The type constructor $\fvar{Arrow}$ is included in the initial constructor context and has index kind $\kpair{\kTypeBlur}{\kTypeBlur}$.
-
-Two type-level terms of kind $\kTypeBlur$ are equivalent if they apply the same constructor, identified by name, to equivalent indices.
-Going further, we ensure that deciding type equivalence requires only checking for syntactic equality after normalization by imposing the restriction that type constructors can only be indexed by kinds for which equivalence can be decided in this way.
-All combinations of kinds introduced thus far  have this property and this is sufficient for many interesting examples.
-Our treatment of equivalence in the type-level language is thus quite similar to the treatment of term-level equality using ``equality types'' in a language like Standard ML.
-An \emph{equality kind} is defined by the judgement \todo{eq type discussion}.
-We discuss introducing a richer notion of  equivalence of types that  fragment providers can extend as future work in Sec.
-\ref{limitations}\todo{do this}.% as maintaining the metatheoretic guarantee that typing respects type equivalence would impose a substantial burden in such a setting.% (a na\"ive approach to this would impose non-trivial extrinsic proof obligations onto extension developers that, unlike in others in this thesis, could threaten type safety).
-
-Type constructors are not first-class; they do not themselves have arrow kind as in some kind systems (e.g.
-\cite{Watkins08}\todo{cite}; Ch.
-22 of \emph{PFPL} describes a related system \cite{pfpl}).
-The type-level language does, however, include total functions of conventional arrow kind, written $\karrow{\kappa_1}{\kappa_2}$.
-Type constructor application can be wrapped in a type-level function to emulate a first-class type constructor (indeed, such a wrapper could be generated automatically, though we avoid this for simplicity in our semantics).
-Equivalence at arrow kind does not meet our criteria, so type-level functions cannot appear within type indices.
-This also prevents general recursion from arising at the type level.
-Without this restriction, a function of type $\karrow{\kTypeBlur}{\kappa}$ could ``smuggle'' in a self-reference as a type index and extract it via case analysis (an issue related to the positivity condition for inductive datatypes in total functional languages).
-
-Every type constructor also defines a \emph{representation schema} using \textsf{schema}, a type-level function that we will discuss in Sec. \ref{?}\todo{ref} after introducing operator constructors.
-
-\subsection{Operator Constructors}
-For reasons that we will discuss, our calculus associates every operator  constructor\footnote{It may be helpful to emphasize that \emph{term formers} are distinct from type constructors and operator constructors.
-There are term formers at all levels in the calculus.
-For example, type constructor application is a type-level term former.
-An abstract syntax at the type level like Harper's \cite{pfpl} might write it $\texttt{tcapp}[\fvar{Tycon}](\tau)$, emphasizing the distinction.} with a type constructor.
-
-User-defined operator constructors are declared using \textsf{opcon} (Fig.
-\ref{kindprog}).
+For reasons that we will discuss, our calculus associates every operator  constructor with a type constructor.
 The \emph{fully-qualified name} of every operator constructor, e.g.
 $\fvar{Nat}.\opvar{z}$, must be unique.
-Operator constructors are indexed by type-level values, like type constructors, and so also declare an index kind $\kappaidx$.
+Operator constructors are indexed by type-level values, like type constructors, and so also declare an index kind, $\kappaidx$.
 In our first example, all the operator constructors are indexed trivially (by index kind $\kunit$), but later examples will use more interesting indices.
-For example, the projection operator for labeled tuples, $\fvar{LabeledTuple}.\opvar{prj}$, has index kind $\klabel$.
-\todo{defne operator}
-Note that neither operator constructors nor operators are first-class type-level values and we do not need to impose any restrictions on their index kinds.
-
-In the external language, an operator is selected and invoked by applying an operator constructor to a type-level index and $n \geq 0$ \emph{arguments}, written $\eop{Tycon}{op}{\tauidx}{\splat{e}{1}{n}}$.
+For example, the projection operator constructor for labeled tuples, $\fvar{LabeledTuple}.\opvar{prj}$, has index kind $\klabel$.
+An operator itself is, notionally, selected by indexing an operator constructor, e.g.
+$\fvar{Nat}.\opvar{s}\langle \tunit \rangle$, but note that neither operator constructors nor operators are first-class at any level (additional machinery would be needed, and this is not fundamental to our calculus).
+
+Instead, in the external language, an operator constructor is applied by simultaneously selecting an operator and invoking it on $n \geq 0$ \emph{arguments}, written $\eop{Tycon}{op}{\tauidx}{\splat{e}{1}{n}}$\footnote{It may be helpful to distinguish between type/operator constructors and \emph{term formers}, which are the metatheoretic entities that ultimately define the syntax.
+There are term formers at all levels in the calculus.
+For example, operator constructor application and $\lambda$ are  external term formers, and type constructor application is a type-level term former.
+An abstract syntax following Harper's conventions highlights this distinction \cite{pfpl}: the term formers $\mathtt{lam}$, $\mathtt{tcapp}$ and $\mathtt{ocapp}$, define, respectively the forms $\mathtt{lam}[\tau](x.e)$, $\texttt{tcapp}[\fvar{Tycon}](\tau)$ and $\mathtt{ocapp}[\fvar{Tycon}, \opvar{op}, \tauidx](\splat{e}{1}{n})$.}.
 For example, on line 18 of Fig.
-\ref{nat}, we see the operator constructors $\fvar{Nat}.\opvar{z}$ and $\fvar{Nat}.\opvar{s}$ being invoked to compute $two$ (the syntactic overhead will be relieved later).
+\ref{nat}, we see the operator constructors $\fvar{Nat}.\opvar{z}$ and $\fvar{Nat}.\opvar{s}$ being applied to compute $two$.
 %\footnote{Although our focus here is entirely on semantics, a brief note on syntax: in the expanded syntax, the trivial indices and empty argument lists can be omitted, so we could write \texttt{Nat.s(Nat.s(Nat.z))}.
@@
 \ref{aparsing}, we might instead just write \texttt{2}.}
-To derive the active typing judgement for this form, the semantics invokes  
- the operator constructor's  \emph{definition}: a type-level function that must examine the provided operator index and the recursively determined \emph{derivates} of the arguments to determine a derivate for the operation as a whole,  or indicate an error.
-A derivate is a type-level representation of the result of deriving the active typing judgement\footnote{Technically, the abstract active typing judgement, which is similar in form and we will introduce shortly.}: an internal term, called the {translation},  paired with a type.
+To derive the active typing judgement for an operator constructor invocation, the semantics calls   
+ into the operator constructor's \emph{definition}: a type-level function that must decide a type assignment and a translation on the basis of the index and arguments, or decide that this is not possible.
+More specifically, in @$\lambda$, examine the provided operator index and the recursively determined \emph{derivates} of the arguments to determine a derivate for the operation as a whole,  or indicate an error.
+A derivate is notionally a type-level representation of the result of deriving the active typing judgement described\footnote{Technically, the abstracted active typing judgement, introduced shortly.}: an internal term, called the {translation},  paired with a type.
 Derivates have kind $\kDen$ and introductory form $\tden{\tautrans}{\tautype}$.
@@
 
-\subsection{Representational Consistency Implies Type Safety}
+\subsection{Representational Consistency Implies Type Safety}\label{repcon}
 
@@
 
-\subsection{Representational Independence Implies Conservativity}
+\subsection{Representational Independence Implies Conservativity}\label{repind}
 In isolation, our definition of natural numbers can be shown to be a strong encoding of the semantics of G\"odel's $\mathbf{T}$.
@@
 
-\section{Semantics and Metatheory}\label{theory}
+\section{Metatheory}\label{theory}
 
@@
 
+The contexts and external term are ``input'' to the judgement, while the type assignment and translation are ``output'' (we briefly discuss potential variants in Sec.
+\ref{limitations}\todo{remove if not}).
 %At best, they can aim to achieve a separation of concerns, such that each component of an application can be written using a specialized language (as advocated by proponents of the \emph{language-oriented approach} to software development \cite{language-oriented}), but there will still be an {interoperability} problem at the interface between components, as described above, unless an FFI has been explicitly developed for each pair of languages that might interface with each other, or library interfaces are constrained to only use a small subset of common constructs (as suggested in draft guidelines for F\#, for example \cite{fsharp-guidelines}) reducing  expressiveness and safety.
@@
 %
-%\text{reified IL}		&		&	\pipe	&	\titerm{\gamma} \pipe \titype{\sigma} \\
+%\text{reified IL}		&		&	\pipe	&	\titerm{\iota} \pipe \titype{\sigma} \\
 %
@@
 %%											 							&		&		&	\\
-%\textbf{internal terms} 				& 	\gamma	&	::=	&	\evar{x} \pipe 
-%												\ilam{\evar{x}}{\sigma}{\gamma} \pipe 
-%												\iapp{\gamma_{1}}{\gamma_{2}} \pipe
-%												\ifix{\evar{f}}{\sigma}{\gamma} \\
+%\textbf{internal terms} 				& 	\iota	&	::=	&	\evar{x} \pipe 
+%												\ilam{\evar{x}}{\sigma}{\iota} \pipe 
+%												\iapp{\iota_{1}}{\iota_{2}} \pipe
+%												\ifix{\evar{f}}{\sigma}{\iota} \\
 %									& & \pipe & 
-%												\ipair{\gamma_{1}}{\gamma_{2}} \pipe 
-%												\ifst{\gamma} \pipe
-%												\isnd{\gamma}  
+%												\ipair{\iota_{1}}{\iota_{2}} \pipe 
+%												\ifst{\iota} \pipe
+%												\isnd{\iota}  
 %												\\
 %							& 		& 	\pipe	& 
-%												\iintlit \pipe \iop{\gamma_{1}}{\gamma_{2}} \pipe \iIfEq{\gamma_{1}}{\gamma_{2}}{\dint}{\gamma_{3}}{\gamma_{4}}  
+%												\iintlit \pipe \iop{\iota_{1}}{\iota_{2}} \pipe \iIfEq{\iota_{1}}{\iota_{2}}{\dint}{\iota_{3}}{\iota_{4}}  
 %												\\
 % &  & \pipe & \tvalof{\tau_1}{\tau_2} \pipe \iup{\tau} \\
-%%\text{deabstracted}& \gamma & ::= & \mathcal{G}[\gamma, \sigma]\\
+%%\text{deabstracted}& \iota & ::= & \mathcal{G}[\iota, \sigma]\\
 %\textbf{internal types}			&	\sigma	&	::=	&    \darrow{\sigma_1}{\sigma_2} \pipe
@@
 Expressions, $e$, can be either variables, lambdas, or applications of operators.
-%%, and are ultimately given meaning by translation to a typed internal language, with terms $\gamma$ and types $\sigma$. 
+%%, and are ultimately given meaning by translation to a typed internal language, with terms $\iota$ and types $\sigma$. 
 %%This language has been chosen, for simplicity, to be a variant of Plotkin's PCF with primitive integers and products, but in practice would include other constructs consistent with its role as a high-level intermediate language.
@@
 This language has been chosen, for simplicity, to be a variant of Plotkin's PCF with primitive integers and products, but in practice would include other constructs consistent with its role as a high-level intermediate language.
-The grammar of internal terms, $\gamma$, and internal types, $\sigma$, also includes special forms containing type-level terms; these are used for developing extensions and during compilation and will be erased before compilation ends.
+The grammar of internal terms, $\iota$, and internal types, $\sigma$, also includes special forms containing type-level terms; these are used for developing extensions and during compilation and will be erased before compilation ends.
 %\subsection{Example: G\"odel's T as an Active Type Family}
@@
 A \emph{valid denotation} has the form $\tden{\tauiterm}{\tautype}$, where $\tauiterm$ is the \emph{translation} of the expression to an internal term and $\tautype$ is the type it has been assigned.
-Internal terms are represented as type-level terms using the form $\titerm{\gamma}$, and have kind $\kITerm$ (similar to $\kIType$).
+Internal terms are represented as type-level terms using the form $\titerm{\iota}$, and have kind $\kITerm$ (similar to $\kIType$).
 %It is important to note that there are no elimination form for reified terms or types.
@@
 It then checks that the argument's type is also $\ttype{Nat}{\tunit}$, returning a denotation pairing the translation $\titerm{\iup{\tvar{x}} + 1}$ with the type $\ttype{Nat}{\tunit}$ if so.
-The form $\iup{\tau}$ is used to ``un-reify'' reified internal terms, of kind $\kITerm$ (thus serving as the left-inverse of $\titerm{\gamma}$).
+The form $\iup{\tau}$ is used to ``un-reify'' reified internal terms, of kind $\kITerm$ (thus serving as the left-inverse of $\titerm{\iota}$).
 In this case, $\tvar{x}$ is a type-level variable of kind $\kITerm$ representing the translation of the argument to the successor operator, so we simply need to add one to it.
@@
 %	\overbrace{\progOK{\emptyctx}{\fvalCtx_0}{\rho}}^{\text{\normalsize Kind Checking}}\\
-%	\overbrace{\pcompiles{\fvalCtx_0}{\rho}{\gamma}}^{\text{\normalsize Active Typechecking and Translation}}
-%}{
-%	\ptcc{\rho}{\gamma}
+%	\overbrace{\pcompiles{\fvalCtx_0}{\rho}{\iota}}^{\text{\normalsize Active Typechecking and Translation}}
+%}{
+%	\ptcc{\rho}{\iota}
 %}
@@
 %To typecheck and translate an expression, $e$, the rule \textsc{att-exp} first assigns a type, $\tau$, and an \emph{abstract translation}, $\gabs$, to it, as determined by the judgement $\ecompilesX{e}{\tau}{\gabs}$.
-It then \emph{deabstracts} this abstract translation to complete the compilation process, as determined by the judgement $\eraseX{\gabs}{\gamma}$.
+It then \emph{deabstracts} this abstract translation to complete the compilation process, as determined by the judgement $\eraseX{\gabs}{\iota}$.
 The purpose of the abstract translation phase is to ensure that the implementation details of type families are not exposed to other families, so that invariants that they rely on are preserved when families are composed.
@@
 %Let us first review how expressions are assigned types and abstract translations.
-Variables (\textsc{att-var}) translate directly to variables in the internal language, with the type determined by the typing context, $\Gamma$.
+Variables (\textsc{att-var}) translate directly to variables in the internal language, with the type determined by the typing context, $\iota$.
 Lambda terms (\textsc{att-lam}) also translate to lambda terms in the internal language and are given the appropriate $\fvar{Arrow}$ type by extending the typing context and proceeding recursively as is usual.
@@
 The premise $\ddbar{\Xi_0,\fvar{Fam}}{\fvalCtx}{\trepof{\tau}}{\sabs}$ determines the abstract internal type and the premise $\checkRC{\gtCtx}{\Xi_0,\fvar{Fam}}{\fvalCtx}{\gabs}{\sabs}$ performs the internal type checking.
-To do so, the variable context, $\Gamma$, which associates variables with types, must be converted to a context associating those variables with their corresponding abstract internal types, $\gtCtx$.
+To do so, the variable context, $\iota$, which associates variables with types, must be converted to a context associating those variables with their corresponding abstract internal types, $\gtCtx$.
 The relevant judgements are defined in Fig.
@@
 %
-%\begin{figure}[t]
-%\small
-%%\vspace{-15pt}
-%$\fbox{\inferrule{}{\tKindX{\tau}{\kappa}}}$
-%\begin{mathpar}
-%\small
-%\inferrule[var-kind]{
-%}{
-%  \tKind{\tvarCtxX{t}{\kappa}}{\fvalCtx}{\tvar{t}}{\kappa}
-%}
-%
-%\inferrule[k-arrow-intro]{
-%  \tKind{\tvarCtxX{t}{\kappa_1}}{\fvalCtx}{\tau}{\kappa_2}
-%}{
-%  \tKindX{\tlam{t}{\kappa_1}{\tau}}{\karrow{\kappa_1}{\kappa_2}}
-%}
-%
-%\inferrule[k-arrow-elim]{
-%  \tKindX{\tau_1}{\karrow{\kappa_1}{\kappa_2}}\\
-%  \tKindX{\tau_2}{\kappa_1}
-%}{
-%  \tKindX{\tapp{\tau_1}{\tau_2}}{\kappa_2}
-%}
-%\\
-%\text{\color{gray} (standard statics for lists, integers, strings and products  omitted)}
-%\\
-%%%
-%%%\inferrule{ }{
-%%%	\tKindX{\tstr{str}}{\kstr}
-%%%}(\text{str}^I_\tau)
-%%%
-%%%\inferrule{ }{
-%%%	\tKindX{\tunit}{\kunit}
-%%%}(\text{1}^I_\tau)
-%%%
-%%%\inferrule{
-%%%	\tKindX{\tau_1}{\kappa_1}\\
-%%%	\tKindX{\tau_2}{\kappa_2}
-%%%}{
-%%%	\tKindX{\tpair{\tau_1}{\tau_2}}{\kpair{\kappa_1}{\kappa_2}}
-%%%}({\times}^I_\tau)
-%%%
-%%%\inferrule{
-%%%	\tKindX{\tau}{\kpair}{\kappa_1}{\kappa_2}
-%%%}{
-%%%	\tKindX{\tfst{\tau}}{\kappa_1}
-%%%}({\times}^{E1}_\tau)
-%%%
-%%%\inferrule{
-%%%	\tKindX{\tau}{\kpair}{\kappa_1}{\kappa_2}
-%%%}{
-%%%	\tKindX{\tsnd{\tau}}{\kappa_2}
-%%%}({\times}^{E2}_\tau)
-%%%\inferrule{ }{
-%%%\inferrule{
-%%%	\tKindX{\tau_{1}}{\kappa_{1}}\\
-%%%	\tKindX{\tau_{2}}{\kappa_{2}}
-%%%}{
-%%%	\tKindX{\tpair{\tau_{1}}{\tau_{2}}}{\kpair{\kappa_{1}}{\kappa_{2}}}
-%%%}~(\times_\tau)
-%%%
-%%%\inferrule{
-%%%	\tKindX{\tau}{\kpair{\kappa_{1}}{\kappa_{2}}}
-%%%}{
-%%%	\tKindX{\tfst{\tau}}{\kappa_{1}}
-%%%}~(\text{fst}_\tau)
-%%%
-%%%\inferrule{
-%%%	\tKindX{\tau}{\kpair{\kappa_{1}}{\kappa_{2}}}
-%%%}{
-%%%	\tKindX{\tsnd{\tau}}{\kappa_{2}}
-%%%}~(\text{snd}_\tau)
-%%%
-%%% TODO: list
-%\inferrule[tl-equality]{
-%	\kEq{\kappa}\\
-%	\tKindX{\tau_1}{\kappa}\\
-%	\tKindX{\tau_2}{\kappa}\\\\
-%	\tKindX{\tau_3}{\kappa'}\\
-%	\tKindX{\tau_4}{\kappa'}
-%}{
-%	\tKindX{\tifeq{\tau_1}{\tau_2}{\kappa}{\tau_3}{\tau_4}}{\kappa'}
-%}
-%
-%\inferrule[type-intro]{
-%	\fvarOfTypeDf \in \fvalCtx\\\\
-%	\tKind{\tvarCtx}{\fvalCtx}{\tauidx}{\kappaidx}
-%}{
-%	\tKindX{\ttype{Fam}{\tauidx}}{\kTypeBlur}
-%}
-%
-%\inferrule[type-elim]{
-%	\fvarOfTypeDf \in \fvalCtx\\
-%	\tKindX{\tau}{\kTypeBlur}\\\\
-%	\tKind{\tvarCtxX{x}{\kappaidx}}{\fvalCtx}{\tau_0}{\kappa}\\
-%	\tKindX{\tau_1}{\kappa}
-%}{
-%	\tKind{\tvarCtx}{\fvalCtx}{\tfamcase{\tau}{Fam}{x}{\tau_0}{\tau_1}}{\kappa}
-%}
-%~~~~~~
-%\inferrule[den-intro-valid]{
-%	\tKindX{\tau_1}{\kITerm}\\\\
-%	\tKindX{\tau_2}{\kTypeBlur}
-%}{
-%	\tKindX{\tden{\tau_1}{\tau_2}}{\kDen}
-%}
-%
-%\inferrule[den-intro-err]{ }{
-%	\tKindX{\terr}{\kDen}
-%}
-%
-%\inferrule[den-elim]{
-%	\tKindX{\tau}{\kDen}\\\\
-%	\tKind{\tvarCtx, \tOfKind{\tvar{x}}{\kITerm}, \tOfKind{\tvar{t}}{\kTypeBlur}}{\fvalCtx}{\tau_1}{\kappa}\\
-%	\tKindX{\tau_2}{\kappa}
-%}{
-%	\tKindX{\tdencase{\tau}{x}{t}{\tau_1}{\tau_2}}{\kappa}
-%}
-%
-%\inferrule[iterm-intro]{
-%	\isIterm{\tvarCtx}{\emptyctx}{\fvalCtx}{\gamma}
-%}{
-%	\tKindX{\titerm{\gamma}}{\kITerm}
-%}
-%
-%\inferrule[itype intro]{
-%	\isItype{\tvarCtx}{\fvalCtx}{\sigma}
-%}{
-%	\tKindX{\titype{\sigma}}{\kIType}
-%}
-%\end{mathpar}
-%$\fbox{$\kEq{\kappa}$}$
-%\begin{mathpar}
-%\inferrule[t-eq]{ }{
-%	\kEq{\kTypeBlur}
-%}
-%
-%\inferrule[z-eq]{ }{
-%	\kEq{\dint}
-%}
-%
-%\inferrule[s-eq]{ }{
-%	\kEq{\kstr}
-%}
-%
-%\inferrule[u-eq]{ }{
-%	\kEq{\kunit}
-%}
-%
-%\inferrule[p-eq]{
-%	\kEq{\kappa_1}\\\\
-%	\kEq{\kappa_2}
-%}{
-%	\kEq{\kpair{\kappa_1}{\kappa_2}}
-%}
-%
-%\inferrule[l-eq]{
-%	\kEq{\kappa}
-%}{
-%	\kEq{\klist{\kappa}}
-%}
-%\end{mathpar}
-%$\fbox{$\isItermX{\gamma}$}$
-%\begin{mathpar}
-%\inferrule[i-var-kinding]{ }{
-%	\isIterm{\tvarCtx}{\eivarCtxX{x}}{\fvalCtx}{\evar{x}}
-%}
-%
-%\inferrule[i-lam-kinding]{
-%	\isItypeX{\sigma}\\\\
-%	\isIterm{\tvarCtx}{\eivarCtxX{x}}{\fvalCtx}{\gamma}
-%}{
-%	\isItermX{\ilam{\evar{x}}{\sigma}{\gamma}}
-%}
-%
-%\inferrule[i-fix-kinding]{
-%	\isItypeX{\sigma}\\\\
-%	\isIterm{\tvarCtx}{\eivarCtxX{x}}{\fvalCtx}{\gamma}
-%}{
-%	\isItermX{\ifix{\evar{f}}{\sigma}{\gamma}}
-%}
-%\\
-%\text{\color{gray} (omitted forms have trivially recursive rules)}\\\vspace{-3pt}
-%\\
-%\inferrule[iterm-dereify-kinding]{
-%	\tKindX{\tau}{\kITerm}
-%}{
-%	\isItermX{\iup{\tau}}
-%}
-%
-%\inferrule[abs-trans-kinding]{
-%	\tKindX{\tauiterm}{\kITerm}\\
-%	\tKindX{\tautype}{\kTypeBlur}
-%}{
-%	\isItermX{\tvalof{\tauiterm}{\tautype}}
-%}
-%\end{mathpar}
-%$\fbox{$\isItypeX{\sigma}$}$
-%\begin{mathpar}
-%\inferrule[i-int-kinding]{ }{
-%	\isItypeX{\dint}
-%}
-%
-%\inferrule[i-prod-kinding]{
-%	\isItypeX{\sigma_1}\\\\
-%	\isItypeX{\sigma_2}
-%}{
-%	\isItypeX{\dpair{\sigma_1}{\sigma_2}}
-%}
-%
-%\inferrule[i-arrow-kinding]{
-%	\isItypeX{\sigma_1}\\\\
-%	\isItypeX{\sigma_2}
-%}{
-%	\isItypeX{\darrow{\sigma_1}{\sigma_2}}
-%}
-%
-%\inferrule[itype-dereify-kinding]{
-%	\tKindX{\tau}{\kIType}
-%}{
-%	\isItypeX{\dup{\tau}}
-%}
-%
-%\inferrule[abs-rep-kinding]{
-%	\tKindX{\tau}{\kTypeBlur}
-%}{
-%	\isItypeX{\trepof{\tau}}
-%}
-%\end{mathpar}
-%\caption{\small Kinding for type-level terms}
-%\label{tlkind}
-%\end{figure}
-%
-%\begin{figure}
-%\small
-%$\fbox{$\tEvalX{\tau}{\tau'}$}$
-%\begin{mathpar}
-%\inferrule[tl-lam-eval]{ }{
-%	\tEvalX{\tlam{t}{\kappa}{\tau}}{\tlam{t}{\kappa}{\tau}}
-%}
-%
-%\inferrule[tl-ap-eval]{
-%	\tEvalX{\tau_1}{\tlam{t}{\kappa}{\tau}}\\
-%	\tEvalX{\tau_2}{\tau_2'}\\
-%	\tEvalX{\subst{\tau_2'}{\tvar{t}}{\tau}}{\tau'}
-%}{
-%	\tEvalX{\tapp{\tau_1}{\tau_2}}{\tau'}
-%}
-%
-%\text{\color{gray} (standard evaluation rules for integers, strings, products and lists omitted)}
-%
-%\inferrule[tl-eq-eval-equal]{
-%	\tEvalX{\tau_1}{\tau_1'}\\
-%	\tEvalX{\tau_2}{\tau_1'}\\
-%	\tEvalX{\tau_3}{\tau_3'}
-%}{
-%	\tEvalX{\tifeq{\tau_1}{\tau_2}{\kappa}{\tau_3}{\tau_4}}{\tau_3'}
-%}
-%
-%\inferrule[tl-eq-eval-inequal]{
-%	\tEvalX{\tau_1}{\tau_1'}\\
-%	\tEvalX{\tau_2}{\tau_2'}\\
-%	\tau_1' \neq \tau_2'\\
-%	\tEvalX{\tau_4}{\tau_4'}
-%}{
-%	\tEvalX{\tifeq{\tau_1}{\tau_2}{\kappa}{\tau_3}{\tau_4}}{\tau_4'}
-%}
-%
-%\inferrule[type-eval]{
-%	\tEvalX{\tauidx}{\tauidx'}
-%}{
-%	\tEvalX{\ttype{Fam}{\tauidx}}{\ttype{Fam}{\tauidx'}}
-%}
-%
-%\inferrule[famcase-eval-match]{
-%	\tEvalX{\tau}{\ttype{Fam}{\tauidx}}\\
-%	\tEvalX{\subst{\tauidx}{\tvar{x}}{\tau_1}}{\tau_1'}
-%}{
-%	\tEvalX{\tfamcase{\tau}{Fam}{x}{\tau_1}{\tau_2}}{\tau_1'}
-%}
-%
-%\inferrule[famcase-eval-fail]{
-%	\tEvalX{\tau}{\ttype{Fam'}{\tauidx}}\\
-%	\fvar{Fam} \neq \fvar{Fam'}\\
-%	\tEvalX{\tau_2}{\tau_2'}
-%}{
-%	\tEvalX{\tfamcase{\tau}{Fam}{x}{\tau_1}{\tau_2}}{\tau_2'}
-%}
-%
-%\inferrule[den-valid-eval]{
-%	\tEvalX{\tau_1}{\tau_1'}\\
-%	\tEvalX{\tau_2}{\tau_2'}
-%}{
-%	\tEvalX{\tden{\tau_1}{\tau_2}}{\tden{\tau_1'}{\tau_2'}}
-%}
-%
-%\inferrule[den-err-eval]{ }{
-%	\tEvalX{\terr}{\terr}
-%}
-%
-%\inferrule[dencase-eval-valid]{
-%	\tEvalX{\tau}{\tden{\tauiterm}{\tautype}}\\\\
-%	\tEvalX{[\titerm{\tvalof{\tauiterm}{\tautype}}/\tvar{x}, \tautype/\tvar{t}]\tau_1}{\tau_1'}
-%}{
-%	\tEvalX{\tdencase{\tau}{x}{t}{\tau_1}{\tau_2}}{\tau_1'}
-%}
-%
-%\inferrule[dencase-eval-err]{
-%	\tEvalX{\tau}{\terr}\\
-%	\tEvalX{\tau_2}{\tau_2'}
-%}{
-%	\tEvalX{\tdencase{\tauden}{y}{x}{\tau_1}{\tau_2}}{\tau_2'}
-%}
-%
-%\inferrule[iterm-reify]{
-%	\tiEvalX{\gamma}{\gamma'}
-%}{
-%	\tEvalX{\titerm{\gamma}}{\titerm{\gamma'}}
-%}
-%
-%\inferrule[itype-reify]{
-%	\tiEvalX{\sigma}{\sigma'}
-%}{
-%	\tEvalX{\titype{\sigma}}{\titype{\sigma'}}
-%}
-%\end{mathpar}
-%$\fbox{$\tiEvalX{\gamma}{\gamma'}$}$
-%\begin{mathpar}
-%\inferrule[i-var-eval]{ }{
-%	\tiEvalX{\evar{x}}{\evar{x}}
-%}
-%
-%\inferrule[i-lam-eval]{
-%	\tiEvalX{\sigma}{\sigma'}\\
-%	\tiEvalX{\gamma}{\gamma'}
-%}{
-%	\tiEvalX{\ilam{\evar{x}}{\sigma}{\gamma}}{\ilam{\evar{x}}{\sigma'}{\gamma'}}
-%}
-%
-%\inferrule[i-fix-eval]{
-%	\tiEvalX{\sigma}{\sigma'}\\
-%	\tiEvalX{\gamma}{\gamma'}
-%}{
-%	\tiEvalX{\ifix{\evar{f}}{\sigma}{\gamma}}{\ifix{\evar{f}}{\sigma'}{\gamma'}}
-%}
-%\\
-%\text{\color{gray} (omitted forms have trivially recursive rules)}
-%\\
-%\inferrule[iterm unquote eval]{
-%	\tEvalX{\tau}{\tau'}
-%}{
-%	\tiEvalX{\iup{\tau}}{\iup{\tau'}}
-%}
-%
-%\inferrule[val from den eval]{
-%	\tEvalX{\tauiterm}{\tauiterm'}\\
-%	\tEvalX{\tautype}{\tautype'}
-%}{
-%	\tiEvalX{\tvalof{\tauiterm}{\tautype}}{\tvalof{\tauiterm'}{\tautype'}}
-%}
-%\end{mathpar}
-%$\fbox{$\tiEvalX{\sigma}{\sigma'}$}$
-%\begin{mathpar}
-%\inferrule[i-int-eval]{ }{
-%	\tiEvalX{\dint}{\dint}
-%}
-%
-%\inferrule[i-prod-eval]{
-%	\tiEvalX{\sigma_1}{\sigma_1'}\\
-%	\tiEvalX{\sigma_2}{\sigma_2'}
-%}{
-%	\tiEvalX{\dpair{\sigma_1}{\sigma_2}}{\dpair{\sigma_1'}{\sigma_2'}}
-%}
-%
-%\inferrule[i-arrow-eval]{
-%	\tiEvalX{\sigma_1}{\sigma_1'}\\
-%	\tiEvalX{\sigma_2}{\sigma_2'}
-%}{
-%	\tiEvalX{\darrow{\sigma_1}{\sigma_2}}{\darrow{\sigma_1'}{\sigma_2'}}
-%}
-%\\
-%\inferrule[itype-eval]{
-%	\tEvalX{\tau}{\tau'}
-%}{
-%	\tiEvalX{\dup{\tau}}{\dup{\tau'}}
-%}
-%
-%\inferrule[abs-rep-eval]{
-%	\tEvalX{\tau}{\tau'}
-%}{
-%	\tiEvalX{\trepof{\tau}}{\trepof{\tau'}}
-%}
-%\end{mathpar}
-%\caption{\small Evaluation semantics for type-level terms}
-%\label{tleval}
-%\end{figure}
+
+%
+
 %
@@
 %\end{mathpar}
-%$\fbox{\inferrule{}{\checkRCX{\gamma}{\sigma}}}$
+%$\fbox{\inferrule{}{\checkRCX{\iota}{\sigma}}}$
 %\begin{mathpar}
@@
 %	\ddbarX{\sigma_1}{\sigma_1'}\\
-%	\checkRC{\gtCtxX{x}{\sigma_1'}}{\Xi}{\fvalCtx}{\gamma}{\sigma_2}
-%}{
-%	\checkRCX{\ilam{x}{\sigma_1}{\gamma}}{\darrow{\sigma_1'}{\sigma_2}}
+%	\checkRC{\gtCtxX{x}{\sigma_1'}}{\Xi}{\fvalCtx}{\iota}{\sigma_2}
+%}{
+%	\checkRCX{\ilam{x}{\sigma_1}{\iota}}{\darrow{\sigma_1'}{\sigma_2}}
 %}
@@
 %\inferrule[abs-i-ap]{
-%	\checkRCX{\gamma_1}{\darrow{\sigma_1}{\sigma_2}}\\
-%	\checkRCX{\gamma_2}{\sigma_1}
-%}{
-%	\checkRCX{\iapp{\gamma_1}{\gamma_2}}{\sigma_2}
+%	\checkRCX{\iota_1}{\darrow{\sigma_1}{\sigma_2}}\\
+%	\checkRCX{\iota_2}{\sigma_1}
+%}{
+%	\checkRCX{\iapp{\iota_1}{\iota_2}}{\sigma_2}
 %}
@@
 %	\ddbarX{\sigma}{\sigma'}\\
-%	\checkRC{\gtCtxX{x}{\sigma'}}{\Xi}{\fvalCtx}{\gamma}{\sigma'}
+%	\checkRC{\gtCtxX{x}{\sigma'}}{\Xi}{\fvalCtx}{\iota}{\sigma'}
 %}{
@@
 %%\inferrule[abs op]{
-%%	\checkRCX{\gamma_1}{\dint}\\
-%%	\checkRCX{\gamma_2}{\dint}
+%%	\checkRCX{\iota_1}{\dint}\\
+%%	\checkRCX{\iota_2}{\dint}
 %%}{
-%%	\checkRCX{\iop{\gamma_1}{\gamma_2}}{\dint}
+%%	\checkRCX{\iop{\iota_1}{\iota_2}}{\dint}
 %%}
@@
 %%\inferrule[abs pair]{
-%%	\checkRCX{\gamma_1}{\sigma_1}\\
-%%	\checkRCX{\gamma_2}{\sigma_2}
+%%	\checkRCX{\iota_1}{\sigma_1}\\
+%%	\checkRCX{\iota_2}{\sigma_2}
 %%}{
-%%	\checkRCX{\ipair{\gamma_1}{\gamma_2}}{\dpair{\sigma_1}{\sigma_2}}
+%%	\checkRCX{\ipair{\iota_1}{\iota_2}}{\dpair{\sigma_1}{\sigma_2}}
 %%}
@@
 %%\inferrule[abs fst]{
-%%	\checkRCX{\gamma}{\dpair{\sigma_1}{\sigma_2}}
+%%	\checkRCX{\iota}{\dpair{\sigma_1}{\sigma_2}}
 %%}{
-%%	\checkRCX{\ifst{\gamma}}{\sigma_1}
+%%	\checkRCX{\ifst{\iota}}{\sigma_1}
 %%}
@@
 %%\inferrule[abs snd]{
-%%	\checkRCX{\gamma}{\dpair{\sigma_1}{\sigma_2}}
+%%	\checkRCX{\iota}{\dpair{\sigma_1}{\sigma_2}}
 %%}{
-%%	\checkRCX{\isnd{\gamma}}{\sigma_2}
+%%	\checkRCX{\isnd{\iota}}{\sigma_2}
 %%}
@@
 %% \inferrule[abs-int-eq]{
-%% 	\checkRCX{\gamma_1}{\dint}\\
-%% 	\checkRCX{\gamma_2}{\dint}\\\\
-%% 	\checkRCX{\gamma_3}{\sigma}\\
-%% 	\checkRCX{\gamma_4}{\sigma}
+%% 	\checkRCX{\iota_1}{\dint}\\
+%% 	\checkRCX{\iota_2}{\dint}\\\\
+%% 	\checkRCX{\iota_3}{\sigma}\\
+%% 	\checkRCX{\iota_4}{\sigma}
 %% }{
-%% 	\checkRCX{\iIfEq{\gamma_1}{\gamma_2}{\dint}{\gamma_3}{\gamma_4}}{\sigma}
+%% 	\checkRCX{\iIfEq{\iota_1}{\iota_2}{\dint}{\iota_3}{\iota_4}}{\sigma}
 %% }
 %\inferrule[abs-iterm-inverse]{
-%	\checkRCX{\gamma}{\sigma}
-%}{
-%	\checkRCX{\iup{\titerm{\gamma}}}{\sigma}
+%	\checkRCX{\iota}{\sigma}
+%}{
+%	\checkRCX{\iup{\titerm{\iota}}}{\sigma}
 %}
@@
 %	\concrep{\fvalCtx}{\ttype{Fam}{\tauidx}}{\sigma}\\
-%	\checkRCX{\gamma}{\sigma}
-%}{
-%	\checkRCX{\tvalof{\titerm{\gamma}}{\ttype{Fam}{\tauidx}}}{\sigma}
+%	\checkRCX{\iota}{\sigma}
+%}{
+%	\checkRCX{\tvalof{\titerm{\iota}}{\ttype{Fam}{\tauidx}}}{\sigma}
 %}
@@
 %	\concrep{\fvalCtx}{\ttype{Fam}{\tauidx}}{\sigma}\\
-%	\checkRCX{\gamma}{\sigma}
-%}{
-%	\checkRCX{\tvalof{\titerm{\gamma}}{\ttype{Fam}{\tauidx}}}{\trepof{\ttype{Fam}{\tauidx}}}
+%	\checkRCX{\iota}{\sigma}
+%}{
+%	\checkRCX{\tvalof{\titerm{\iota}}{\ttype{Fam}{\tauidx}}}{\trepof{\ttype{Fam}{\tauidx}}}
 %}
@@
 %%\begin{theorem}[Representational Consistency]
-%%If $\progOK{ }{\Sigma}{e}$ and $\ecompiles{ }{\fvalCtx}{e}{\tau}{\gabs}$ and $\eraseX{\gabs}{\gamma}$ and 	$\ddbar{\Xi_0}{\fvalCtx}{\trepof{\tau}}{\sabs}$ and $\eraseX{\sabs}{\sigma}$ then $\checkRC{ }{\Xi_0}{\fvalCtx}{\gamma}{\sigma}$.
+%%If $\progOK{ }{\Sigma}{e}$ and $\ecompiles{ }{\fvalCtx}{e}{\tau}{\gabs}$ and $\eraseX{\gabs}{\iota}$ and 	$\ddbar{\Xi_0}{\fvalCtx}{\trepof{\tau}}{\sabs}$ and $\eraseX{\sabs}{\sigma}$ then $\checkRC{ }{\Xi_0}{\fvalCtx}{\iota}{\sigma}$.
 %%\end{theorem}
@@
 %%\begin{theorem}[Non-Interference]
-%%For any property $P(\gamma)$, if $\progOK{ }{\Sigma}{e}$ and $\ecompiles{ }{\fvalCtx}{e}{\tau}{\gabs}$ and $\eraseX{\gabs}{\gamma}$ implies $P(\gamma)$, then for any $\Sigma'$ and $\fvalCtx'$ disjoint from $\Sigma$ and $\fvalCtx$, we have that if $\progOK{ }{\Sigma,\Sigma'}{e'}$ and $\ecompiles{ }{\fvalCtx,\fvalCtx'}{e'}{\tau}{\gabs'}$ and $\eraseX{\gabs'}{\gamma'}$ then $P(\gamma')$.
+%%For any property $P(\iota)$, if $\progOK{ }{\Sigma}{e}$ and $\ecompiles{ }{\fvalCtx}{e}{\tau}{\gabs}$ and $\eraseX{\gabs}{\iota}$ implies $P(\iota)$, then for any $\Sigma'$ and $\fvalCtx'$ disjoint from $\Sigma$ and $\fvalCtx$, we have that if $\progOK{ }{\Sigma,\Sigma'}{e'}$ and $\ecompiles{ }{\fvalCtx,\fvalCtx'}{e'}{\tau}{\gabs'}$ and $\eraseX{\gabs'}{\iota'}$ then $P(\iota')$.
 %%\end{theorem}
@@
 % CANT GUARANTEE THAT SPECIFICATIONS ARE ACTUALLY DECIDABLE
+\begin{figure}
+\small
+$\fbox{$\tEvalX{\tau}{\tau'}$}$
+\begin{mathpar}
+\inferrule[n-t-lam]{ }{
+	\tEvalX{\tlam{t}{\kappa}{\tau}}{\tlam{t}{\kappa}{\tau}}
+}
+
+\inferrule[n-t-ap]{
+	\tEvalX{\tau_1}{\tlam{t}{\kappa}{\tau}}\\
+	\tEvalX{\tau_2}{\tau_2'}\\
+	\tEvalX{\subst{\tau_2'}{\tvar{t}}{\tau}}{\tau'}
+}{
+	\tEvalX{\tapp{\tau_1}{\tau_2}}{\tau'}
+}
+
+\text{\color{gray} (normalization for integers, labels, lists, products and sums also standard)}
+
+\inferrule[n-t-eq-equal]{
+	\tEvalX{\tau_1}{\tau_1'}\\
+	\tEvalX{\tau_2}{\tau_1'}\\
+	\tEvalX{\tau_3}{\tau_3'}
+}{
+	\tEvalX{\tifeq{\tau_1}{\tau_2}{\kappa}{\tau_3}{\tau_4}}{\tau_3'}
+}
+
+\inferrule[n-t-eq-disequal]{
+	\tEvalX{\tau_1}{\tau_1'}\\
+	\tEvalX{\tau_2}{\tau_2'}\\\\
+	\tau_1' \neq \tau_2'\\
+	\tEvalX{\tau_4}{\tau_4'}
+}{
+	\tEvalX{\tifeq{\tau_1}{\tau_2}{\kappa}{\tau_3}{\tau_4}}{\tau_4'}
+}
+
+\inferrule[nt-ty]{
+	\tEvalX{\tauidx}{\tauidx'}
+}{
+	\tEvalX{\ttype{Tycon}{\tauidx}}{\ttype{Tycon}{\tauidx'}}
+}
+
+\inferrule[nt-tycase-match]{
+	\tEvalX{\tau}{\ttype{Tycon}{\tauidx}}\\
+	\tEvalX{\subst{\tauidx}{\tvar{x}}{\tau_1}}{\tau_1'}
+}{
+	\tEvalX{\tfamcase{\tau}{Tycon}{x}{\tau_1}{\tau_2}}{\tau_1'}
+}
+
+\inferrule[nt-tycase-fail]{
+	\tEvalX{\tau}{\ttype{Tycon'}{\tauidx}}\\
+	\fvar{Tycon} \neq \fvar{Tycon'}\\
+	\tEvalX{\tau_2}{\tau_2'}
+}{
+	\tEvalX{\tfamcase{\tau}{Tycon}{x}{\tau_1}{\tau_2}}{\tau_2'}
+}
+
+\inferrule[nt-d]{
+	\tEvalX{\tau}{\tau'}\\
+	\tiEvalX{\ibar}{\ibar'}
+}{
+	\tEvalX{\tden{\ibar}{\tau}}{\tden{\ibar'}{\tau'}}
+}
+
+\inferrule[nt-d-checked]{
+	\tEvalX{\tau}{\tau'}\\
+	\tiEvalX{\ibar}{\ibar'}
+}{
+	\tEvalX{\tden{\ibar}{\tau}^\checkmark}{\tden{\ibar'}{\tau'}^\checkmark}
+}
+
+\inferrule[nt-typeof]{
+	\tEvalX{\tau}{\tden{\ibar}{\tau'}}
+}{
+	\tEvalX{\ttypeof{\tau}}{\tau'}
+}
+
+\inferrule[nt-typeof-checked]{
+	\tEvalX{\tau}{\tden{\ibar}{\tau'}^\checkmark}
+}{
+	\tEvalX{\ttypeof{\tau}}{\tau'}
+}
+
+\inferrule[nt-reptype]{
+	\tiEvalX{\sbar}{\sbar'}
+}{
+	\tEvalX{\titype{\sbar}}{\titype{\sbar'}}
+}
+\end{mathpar}
+$\fbox{$\tiEvalX{\ibar}{\ibar'}$}$
+\begin{mathpar}
+\inferrule[nt-i-var]{ }{
+	\tiEvalX{\evar{x}}{\evar{x}}
+}
+~~~~~
+%\inferrule[nt-i-lam]{
+%	\tiEvalX{\sigma}{\sigma'}\\
+%	\tiEvalX{\iota}{\iota'}
+%}{
+%	\tiEvalX{\ilam{\evar{x}}{\sigma}{\iota}}{\ilam{\evar{x}}{\sigma'}{\iota'}}
+%}
+%
+\inferrule[nt-i-fix]{
+	\tiEvalX{\sbar}{\sbar'}\\
+	\tiEvalX{\ibar}{\ibar'}
+}{
+	\tiEvalX{\ifix{\evar{x}}{\sbar}{\ibar}}{\ifix{\evar{x}}{\sbar'}{\ibar'}}
+}
+~~~~~
+\inferrule[nt-i-transof]{
+	\tEvalX{\tau}{\tau'}
+}{
+	\tiEvalX{\itransof{\tau}}{\itransof{\tau'}}
+}
+\\
+\text{\color{gray} (omitted forms have similarly recursive rules)}
+%\inferrule[iterm unquote eval]{
+%	\tEvalX{\tau}{\tau'}
+%}{
+%	\tiEvalX{\iup{\tau}}{\iup{\tau'}}
+%}
+%
+\end{mathpar}
+$\fbox{$\tiEvalX{\sbar}{\sbar'}$}$
+\begin{mathpar}
+%\inferrule[i-int-eval]{ }{
+%	\tiEvalX{\dint}{\dint}
+%}
+%
+%\inferrule[i-prod-eval]{
+%	\tiEvalX{\sigma_1}{\sigma_1'}\\
+%	\tiEvalX{\sigma_2}{\sigma_2'}
+%}{
+%	\tiEvalX{\dpair{\sigma_1}{\sigma_2}}{\dpair{\sigma_1'}{\sigma_2'}}
+%}
+%
+\inferrule[nt-i-parr]{
+	\tiEvalX{\sbar_1}{\sbar_1'}\\
+	\tiEvalX{\sbar_2}{\sbar_2'}
+}{
+	\tiEvalX{\darrow{\sbar_1}{\sbar_2}}{\darrow{\sbar_1'}{\sbar_2'}}
+}
+
+\inferrule[nt-i-unquote]{
+	\tEvalX{\tau}{\tau'}
+}{
+	\tiEvalX{\dup{\tau}}{\dup{\tau'}}
+}
+
+\inferrule[nt-i-repof]{
+	\tEvalX{\tau}{\tau'}
+}{
+	\tiEvalX{\trepof{\tau}}{\trepof{\tau'}}
+}
+
+\text{\color{gray} (omitted forms have similarly recursive rules)}
+\end{mathpar}
+\caption{\small Evaluation semantics for type-level terms}
+\label{tleval}
+\end{figure}
+
 \bibliographystyle{abbrv}

### 2014-03-13 sciunit-icse14/sciunit-icse14.tex
Updated intro to ICSE14

@@
 Today, a scientist demonstrates the validity of a model by making an argument in a paper and submitting it for peer review, a process comparable to \emph{code review} in software engineering.
-While human review helps to ensure that contributions meet high-level goals, software engineers typically supplement this with \emph{unit testing} to get a more complete picture of the status of a software project, particularly for complex projects involving many developers.
+While human review helps to ensure that contributions meet high-level goals, software engineers typically supplement it with \emph{unit testing} to get a more complete picture of the status of a project, particularly when it is complex and involves many contributors.
 
-We argue that a similar test-driven methodology would be valuable to scientific communities as they seek to  validate increasingly complex models against growing collections of empirical data.
-The dynamics of  scientific communities and software communities differ in several key ways, however.
-In this paper, we introduce \emph{SciUnit}, a framework for test-driven scientific model validation.
-We describe how SciUnit, supported by new and existing collaborative infrastructure, can integrate into the modern scientific process.
+We argue that a similar test-driven methodology would be valuable to scientific communities as they seek to  validate increasingly complex models against growing repositories of empirical data.
+Scientific communities differ from software communities in several key ways, however.
+In this paper, we introduce \emph{SciUnit}, a framework for test-driven scientific model validation and outline how SciUnit, supported by new and existing collaborative infrastructure, could be  integrated into the modern scientific workflow.
 %in the context of the applications and architectures that the language may need to target.
@@
 
-Scientific theories often take the form of a \emph{quantitative model}: a formal structure that can generate predictions about observable quantities.
-Such a model is characterized by its \textit{scope}: the set of observable quantities that the model can predict, and by its \textit{validity}: the extent to which its predictions agree with experimental observations of these quantities.
+Scientific theories are increasingly being organized around \emph{quantitative models}:  formal systems capable of generating predictions about observable quantities.
+A model can be  characterized by its \textit{scope}: the set of observable quantities that it attempts to predict, and by its \textit{validity}: the extent to which its predictions agree with experimental observations of these quantities.
 
-Quantitative models are validated by \emph{peer review}.
-For a model to be accepted by the scientific community, scientists must submit a paper describing it and providing  evidence that it predicts some quantity of interest more accurately than previous models, or that it makes a desirable tradeoff  between accuracy and complexity \cite{box1987empirical}.
-Other members of the relevant community are then tasked with ensuring that validity was measured properly and that relevant data and competing models were adequately considered, drawing on knowledge of statistical methods and the prior literature.
+Quantitative models are today validated by \emph{peer review}.
+For a model to be accepted by a scientific community, its advocates must submit a paper describing how it works and providing evidence that it predicts a quantity of interest more accurately than previous models, or in some cases, that it makes a desirable tradeoff  between accuracy and complexity \cite{box1987empirical}.
+Other members of the relevant community are then tasked with ensuring that validity was measured properly and that relevant data and competing models were adequately considered, drawing on knowledge of statistical methods and on the prior literature.
 Publishing is a primary motivator for most scientists \cite{howison2011scientific}.
 
-Quantitative scientific modeling and software development have much in common.
-Indeed, quantitative models are increasingly  implemented as computer programs and in some cases, the program \emph{is} the model (e.g.
+Quantitative scientific modeling and software development share much in common.
+Indeed, quantitative models are increasingly being implemented in software and in some cases, the software \emph{is} the model (e.g.
 complex simulations).
-The peer review process for papers is similar in many ways to the \emph{code review} process used in many development teams, where team members look for errors, enforce style and architectural guidelines and check that the code is \emph{valid} (i.e.
+The peer review process for papers is similar in many ways to the \emph{code review} process used in many development teams, where team members look for mistakes, enforce style and architectural guidelines and check that the code is \emph{valid} (i.e.
 that it achieves its intended goal) before permitting it to be committed to the primary source code repository.
@@
 Code review can be quite effective \cite{codereview}, but this requires that developers  expend considerable effort \cite{kemerer2009impact}. 
-Code review is also most effective for resolving  issues related to software architecture.
-Most development teams thus supplement code reviews with {automated} approaches to verification and validation, the most widely-used  of which is \emph{unit testing} \cite{beck2003}.
-In brief, unit tests are functions that check a single portion of a program against a single well-defined correctness criterion.
-A suite of such tests can be seen as a partial specification of a program or component.
-Test suites complement code review by allowing developers to answer questions like these more easily:
+%Code review is also most effective for resolving  issues related to software architecture. 
+Most large development teams thus supplement code reviews with more {automated} approaches to verification and validation, the most widely-used  of which is \emph{unit testing} \cite{beck2003}.
+In brief, unit tests are functions that check that the output of a single component satisfies a single well-defined  criterion.
+A suite of such tests can be considered a partial specification of the program or component under development.
+Test suites serve as a complement to code review by allowing developers to answer questions like these more easily:
 \begin{enumerate}
@@
 %\item How much progress does an individual code contribution make? 
-\item What modes of usage does the team consider most important? How does the team measure correctness?
-\item Does a code contribution cause \emph{regressions} in other parts of a program?
+\item How does the team measure validity?
+\item Does a candidate code contribution cause \emph{regressions} in other parts of a program?
 \end{enumerate}
@@
 \begin{enumerate}
-\item Which observations are already well-explained by existing models? What is the state-of-the-art? What are the open modeling problems of interest?
+\item Which observations are already explained by existing models? What are the best models? What are the open modeling problems of interest?
 %\item How much better is a new model than the state of the art? What are the open problems in my field?
@@
 Each publication focuses on just one model and is frozen in time, so it does not consider the latest experimental data or statistical methods.
-Discovering, precisely characterizing and comparing models to discover the state of the art and find open modeling problems  require an encyclopedic knowledge of the literature.
-Senior scientists can attempt to fill this need by publishing review papers, but in many areas, the number of publications generated every year can be overwhelming \cite{jinha_article_2010}, and comprehensive reviews of a particular area are published relatively infrequently.
+Discovering, precisely characterizing and comparing models to discover the state of the art and find open modeling problems can require an encyclopedic knowledge of the literature.
+Senior scientists often attempt to fill this need by publishing review papers, but in many areas, the number of publications generated every year can be overwhelming \cite{jinha_article_2010}, and comprehensive reviews of a particular area are published relatively infrequently.
 Statisticians often complain that scientists are not following best practices and that community standards evolve too slowly because a canonical paper or popular review used outdated methods.
@@
 
-One might compare this to a software development team relying primarily on carefully-prepared and cross-referenced summaries of code review sessions to understand, validate and evolve a complex software project where  portions of the original source code have been obscured or lost.
+One might compare this to a large software development team relying primarily on carefully prepared and reviewed component documentation together with occasional high-level project summaries produced by the lead architects to answer the questions enumerated above.
 Although certainly a caricature, this motivates our suggestion that the scientific process could be improved by the adoption of test-driven methodologies alongside traditional peer review.
-However, the scientific community presents several challenges that must be addressed by any such methodology:
+However, the scientific community presents several unique challenges that must be addressed by any such methodology:
 \begin{enumerate}
@@
 %\item Experimental data can itself be erroneous, noisy or inconsistent with other available data.
-\item Different modelers often  prefer different programming languages or frameworks.
-Similarly, data formats can differ between experimentalists.
-The testing framework must be as flexible as possible about these choices.
+\item Different modelers often  prefer different programming languages and data structures.
+Similarly, data formats differ between experimentalists.
+The validation  testing framework must be flexible about these choices.
 \item Professional software developers are typically trained in testing practices and tools, while scientists  rarely have training or experience with testing practices \cite{oai:open.ac.uk.OAI2:17673}.
@@
 To address this, we introduce a community workflow based on widely-adopted social coding tools (here Github) along with a lightweight community portal called \textit{SciDash}.
-The overall goal of this work is to help scientists generate and examine tables like the one shown to Figure 1, where the relative validity of a set of models having a common scope can be determined by examining scores produced by a suite of validation tests constructed from experimental data.
+The overall goal of this work is to help scientists generate and examine tables like the one central to Figure 1, where the relative validity of a set of models having a common scope can be determined by examining scores produced by a suite of validation tests constructed from experimental data.
 

### 2014-03-13 sciunit-icse14/sciunit-icse14.tex
Changes for reviewers

@@
 Before explaining the details of this example, we point out that SciUnit is implemented in Python.
-Python is one of the most widely used languages in science today \cite{sanner1999python}.
+Python is one of the most widely used languages in science today \cite{sanner1999python}, and has become the de facto language of open source scientific computing.
 It supports calling into  many of the other popular languages, including R, MATLAB, C and Java, more cleanly than these languages support calling into Python (challenge 3).
 It is widely-recognized as being easy-to-read and its object system can be used to define abstract interfaces, which we leverage in support of challenge 2.
+Alternative implementations of SciUnit, including language-independent specifications of its structure, may be part of future work.
 
@@
 
-While we discuss a toy example based on planetary movement here, we have applied this framework to more realistic problems in neurobiology and have bridged SciUnit to large-scale neuroinformatics projects (we anticipate publishing a description of these case studies shortly).
+While we discuss a toy example based on planetary movement here, we have applied this framework to more realistic problems in neurobiology in a project called NeuronUnit (http://github.com/scidash/neuronunit).
+We have used this as part of a collaboration with a large-scale neuroinformatics project (http://www.opensourcebrain.org), whose developers are enthusiastic about a practical pipeline for testing the dozens of models they currently maintain.
 Much future work remains to be done to investigate whether these tools are truly usable and useful to scientific communities, to further develop the community workflow and infrastructure and to develop a range of realistic case studies.
@@
 3.
-%In particular, we consider modeling competitions in science as a form of test-driven validation.
-For example, the quantitative single neuron modeling competition (QSNMC) \cite{jolivet_quantitative_2008} investigates the complexity-accuracy tradeoff among reduced models of excitable membranes, and
-%examples from other subfields of biology abound (e.g.
-\url{http://www.the-dream-project.org}).
-These competitions typically use \emph{ad hoc} infrastructure.
+
+We also plan to use SciUnit to implement competitions among classes of models or techniques in particular subfields.
+Typically such competitions use \emph{ad hoc} infrastructure.
 A common framework could be used to make designing and continuously operating such contests more feasible.
-In related work that
 

### 2014-03-13 sciunit-icse14/sciunit-icse14.tex
More changes for reviewers

@@
 % can use linebreaks \\ within to get better formatting as desired
-\title{Collaborative Infrastructure for Test-Driven\\Scientific Model Validation}
+\title{SciUnit: Collaborative Infrastructure for Test-Driven\\Scientific Model Validation}
 
@@
 The interface between data collection software and existing data representation standards and tests will be mediated by bridge logic also contained in repositories like \verbx{cosmounit}.
+Because Github is already used by millions of developers for collaborative projects, leveraging this infrastructure should facilitate development.
 % This workflow is meant to evoke the \emph{test-driven development} methodology that is often used in software projects.
@@
 by adding the model a scientist is developing to it), a scientist can simply fork the repository.
-SciDash automatically discovers public forks of repositories that it is already indexing.
+SciDash automatically discovers public forks of repositories (on GitHub) that it is already indexing.
 
@@
 
+The core SciUnit framework has been fully developed and is available at http://sciunit.scidash.org/.
+SciDash is under active development, and is currently capable of basic forms of the operations described in Sec.
+3.
 While we discuss a toy example based on planetary movement here, we have applied this framework to more realistic problems in neurobiology in a project called NeuronUnit (http://github.com/scidash/neuronunit).
@@
 Nevertheless, we believe that identifying the synergies between testing practices and model validation, and describing the basic tooling, represents a novel contribution to the literature on software testing.
-The core SciUnit framework has been fully developed and is available at http://sciunit.scidash.org/.
-SciDash is under active development, and is currently capable of basic forms of the operations described in Sec.
-3.
 
+Future domain-specific layers on top of SciUnit, by analogy to NeuronUnit above, can be developed by interested communities.
+Importantly, interfacing models in new fields to SciUnit, requires only exposing the capabilities of those models as described in the text.
 We also plan to use SciUnit to implement competitions among classes of models or techniques in particular subfields.

### 2014-03-13 sciunit-icse14/sciunit-icse14.tex
ICSE14: Added ACM requirements, updated intro and figure caption

--- 2014-03-13 sciunit-icse14/sciunit-icse14.tex
@@
 
+%\usepackage[compact]{titlesec}
+%\titlespacing{\section}{0pt}{*0}{*0}
+
 %\usepackage{biblatex}
@@
 %\usepackage{times}
-\usepackage{hyperref}
+%\usepackage{hyperref}
 %\usepackage{graphicx}
@@
 \renewcommand{\ttdefault}{txtt}
+
+%\titlespacing{\subsection}{0pt}{*0}{*0}
+%\titlespacing{\subsubsection}{0pt}{*0}{*0}
+
 %%\usepackage{color}
@@
 
+\usepackage{enumitem}
+\setenumerate{noitemsep,topsep=2pt,parsep=0pt,partopsep=0pt}
+\newcommand{\subparagraph}{}
+\usepackage[compact]{titlesec}
+\titlespacing{\section}{0pt}{*0}{*0}
+\titlespacing{\subsection}{0pt}{*0}{*0}
+\titlespacing{\subsubsection}{0pt}{*0}{*0}
+
 \begin{document}
+\conferenceinfo{ICSE}{'14, May 31 - June 7, 2014, Hyderabad, India}
+\CopyrightYear{2014}
+\crdata{978-1-4503-2768-8/14/05}
 %
@@
 \author{\alignauthor
-Cyrus Omar, Jonathan Aldrich\\\affaddr{Carnegie Mellon University}\\
-       \email{\{comar,aldrich\}@cs.cmu.edu}
+Cyrus Omar, Jonathan Aldrich\\\affaddr{Carnegie Mellon University, USA}\\
+       {\{comar,aldrich\}@cs.cmu.edu}
        \alignauthor Richard C. Gerkin\\
-       \affaddr{Arizona State University}\\{rgerkin@asu.edu}
+       \affaddr{Arizona State University, USA}\\{rgerkin@asu.edu}
 }
@@
 \end{abstract}
+\category{D.2.5}{Software Engineering}{Testing and Debugging}%[complexity measures, performance measures]
+\terms{Documentation, measurement, standardization}
+\keywords{unit testing, model validation, cyberinfrastructure}
 
@@
 \vspace{-40px}
-\caption{Tests are derived from data and models are derived from scientific theories.
+\caption{\small Tests are derived from data and models are derived from scientific theories.
 The score table summarizes the performance of a collection of models against a suite of tests.
-A table is contained in a \emph{suite repository} (e.g.
-\texttt{saturnsuite}), hosted on social coding infrastructure -- here, Github.
-SciDash is a portal that discovers and organizes these suite repositories  and provides tools for visualizing them.
-Tests interface with models using capabilities, as specified by the testing framework (\texttt{sciunit}).
-Common testing code, capabilities and bridges to external model and data repositories are also collaboratively developed (e.g.
+A table is generated and visualized in an IPython notebook stored inside a \emph{suite repository} (e.g.
+\texttt{saturnsuite}) hosted on social coding infrastructure (here, Github).
+SciDash is a portal that discovers and organizes these suite repositories and provides read-only views of the notebooks they contain.
+Common testing code, capabilities and bridges to external model and data repositories are also collaboratively developed in \emph{common repositories} (e.g.
 \texttt{cosmounit}).}
 \label{fig:sciunit_overview}
+\vspace{-10pt}
 \end{figure*}
@@
 Most large development teams thus supplement code reviews with more {automated} approaches to verification and validation, the most widely-used  of which is \emph{unit testing} \cite{beck2003}.
-In brief, unit tests are functions that check that the output of a single component satisfies a single well-defined  criterion.
-A suite of such tests can be considered a partial specification of the program or component under development.
-Test suites serve as a complement to code review by allowing developers to answer questions like these more easily:
+In brief, unit tests are functions that check that the behavior of a single component satisfies a single functional  criterion.
+%A suite of such tests can be considered a partial specification of the program or component under development.  
+A suite of such tests complements code review by making it easier to answer questions like these more easily:
 \begin{enumerate}
-\item Which functionality has been adequately implemented? What remains to be done?
+\item What functionality should each component have?
+\item What functionality has been adequately implemented? What remains to be done?
 %\item How much progress does an individual code contribution make? 
-\item How does the team measure validity?
 \item Does a candidate code contribution cause \emph{regressions} in other parts of a program?
@@
 \begin{enumerate}
+\item What are the contemporary community standards for measuring goodness-of-fit?
 \item Which observations are already explained by existing models? What are the best models? What are the open modeling problems of interest?
 %\item How much better is a new model than the state of the art? What are the open problems in my field?
-\item What are the contemporary community standards for measuring goodness-of-fit?
-\item How do newly-made experimental observations impact the validity of previously-published models?
+\item How do newly-made experimental observations impact the validity of previously-published models? Can new models explain previously published data?
 \end{enumerate}
@@
 Each publication focuses on just one model and is frozen in time, so it does not consider the latest experimental data or statistical methods.
-Discovering, precisely characterizing and comparing models to discover the state of the art and find open modeling problems can require an encyclopedic knowledge of the literature.
+Discovering, precisely characterizing and comparing models to discover the state of the art and find open modeling problems can require an encyclopedic knowledge of the literature, as can finding all data relevant to a model.
 Senior scientists often attempt to fill this need by publishing review papers, but in many areas, the number of publications generated every year can be overwhelming \cite{jinha_article_2010}, and comprehensive reviews of a particular area are published relatively infrequently.
 Statisticians often complain that scientists are not following best practices and that community standards evolve too slowly because a canonical paper or popular review used outdated methods.
-Furthermore, if the literature simply doesn't address an important question of validity, a researcher often needs to reimplement a model from scratch.
+Furthermore, if the literature simply doesn't address an important question of validity, because  new data has been gathered since the model paper was published for example, a researcher might need to reimplement an existing model in order to evaluate it.
 
-One might compare this to a large software development team relying primarily on carefully prepared and reviewed component documentation together with occasional high-level project summaries produced by the lead architects to answer the questions enumerated above.
+One might compare this to a large software development team answering the questions listed above based only on carefully reviewed component documentation together with annual high-level project summaries.
 Although certainly a caricature, this motivates our suggestion that the scientific process could be improved by the adoption of test-driven methodologies alongside traditional peer review.
@@
 \item Unit tests are typically pass/fail, while goodness-of-fit between a model and data is typically measured by a continuous metric (e.g. a $p$-value). 
-\item Unit tests often test a \emph{particular} component, whereas a \emph{validation test} must be able to uniformly handle any model capable of predicting the quantity being tested.
+\item Unit tests often test a \emph{particular} component, whereas a \emph{validation test} must be able to handle and compare many models capable of predicting the same quantity.
 %They are sometimes reluctant to share certain details of their implementations \cite{badscientists}.
 %\item Experimental data can itself be erroneous, noisy or inconsistent with other available data.
-\item Different modelers often  prefer different programming languages and data structures.
-Similarly, data formats differ between experimentalists.
-The validation  testing framework must be flexible about these choices.
+\item Different modelers use different programming languages.
 \item Professional software developers are typically trained in testing practices and tools, while scientists  rarely have training or experience with testing practices \cite{oai:open.ac.uk.OAI2:17673}.
 Thus, the framework must be as simple as possible.
-\item Different communities, groups and individuals prefer different goodness-of-fit metrics and emphasize different sets of observable quantities.
-These can evolve as, for example, statisticians develop better measures.
-In contrast, there is more pressure to agree upon requirements and priorities in a software development project.
+\item Different communities, groups and individuals prefer different goodness-of-fit metrics and care about different sets of observable quantities.
+In contrast, there is typically more pressure to agree upon requirements and priorities in a software development project.
 \end{enumerate}
@@
   h, p = randomly_chunk(saturn_obs_positions)
-  saturn_position_test = SpikeCountTest(h, p)
+  saturn_position_test = PositionTest(h, p)
 \end{python}

### 2014-03-13 sciunit-icse14/sciunit-icse14.tex
Code updated to match SciUnit

--- 2014-03-13 sciunit-icse14/sciunit-icse14.tex
@@
     Parameters: 
-      obs_histories : list[list[position]]
-      obs_positions: list[position]"""
-	def __init__(self, obs_histories, obs_positions):
-		self.obs_histories = obs_histories
-		self.obs_positions = obs_positions
+      observation : ['histories':histories, 'positions':positions]
+        histories : list[list[position]]
+        positions : list[position]"""
 
@@
 
-	def _judge(self, model):
+	def generate_prediction(self, model):
 		predictions = []
-		for obs_history in self.obs_histories:
+		for obs_history in self.observation['histories']:
 			predictions.append(model.predict_next_pos(obs_history))
-		p = pooled_p_val(predictions, self.obs_positions)
-		return sciunit.PValue(p, related_data={
-			'obs_histories': obs_histories,
+		return predictions
+
+	def compute_score(observation, prediction):
+		obs_positions = observation['positions']
+		p = pooled_p_val(predictions, obs_positions)
+		return sciunit.scores.PValue(p, related_data={
+			'obs_histories': observation['histories'],
 			'obs_positions': obs_positions,
@@
 A \textit{SciUnit} validation test is an {instance} of a Python class implementing the \verbx{sciunit.Test} abstract interface (line 1). 
-Here, the class \verbx{PositionTest} takes two \emph{parameters} in its constructor (constructors are named \verbx{\_\_init\_\_} in Python, lines 7-9). 
-The meaning of each parameter along with a description of the goodness-of-fit metric used by the test is documented on lines 2-6. 
+A test takes one required parameter in its constructor (constructors are named \verbx{\_\_init\_\_} in Python, lines 7-9), summarizing observed data against which the test will validate models. 
+For the class \verbx{PositionTest}, this observation parameter is a dictionary containing two kinds of observations.   
+The meaning of this parameter along with a description of the goodness-of-fit metric used by the test is documented on lines 2-6. 
 %For convenience, we also make use of functions provided by the popular NumPy\cite{numpy_url} and SciPy\cite{scipy_url} libraries, although these are not required by \textit{SciUnit}.  
@@
   h, p = randomly_chunk(saturn_obs_positions)
-  saturn_position_test = SpikeCountTest(h, p)
+  saturn_position_test = SpikeCountTest({'histories':h, 'positions':p)
 \end{python}
@@
 
-Classes that implement the \verbx{sciunit.Test} interface must contain a \verbx{_judge} method that receives a candidate \emph{model} as input and produces a \textit{score} as output. 
+Classes that implement the \verbx{sciunit.Test} interface must contain a \verbx{generate_prediction} method that receives a candidate \emph{model} as input and produces a prediction as output. 
+They must also contain a \verbx{compute_score} method that can take this prediction, and the observation used to instantiate the test, and produce a \verbx{score}.  
 To specify the interface between the test and the model, the test author provides a list of \emph{capabilities} in the \verbx{required_capabilities} attribute, seen on line 11 of Fig.
@@
 
-The remainder of the \verbx{_judge} method compares the model predictions to observed data to produce a pooled $p$ value.
-The method returns an instance of \verbx{sciunit.PValue}, a subclass of \texttt{sciunit.Score} that has been included with \textit{SciUnit} due to its generality.
+The remainder of the \verbx{compute_score} method compares the model predictions to observed data to produce a pooled $p$ value.
+The method returns an instance of \verbx{sciunit.scores.PValue}, a subclass of \texttt{sciunit.Score} that has been included with \textit{SciUnit} due to its generality.
 In addition to the $p$-value itself, the returned score object also contains metadata, via the \verbx{related_data} parameter, for  scientists who may wish to examine the result in more detail later.
@@
 
-This method proceeds by first checking that the provided model implements all required capabilities before calling the test's \verbx{_judge} method to produce a score.
+This method proceeds by first checking that the provided model implements all required capabilities before calling the test's \verbx{compute_score} method to produce a score.
 A reference to the test and model are added to the score for convenience (accessible via the \verbx{test} and \verbx{model} attributes, respectively).

### 2014-03-14 sciunit-icse14/sciunit-icse14.tex
ICSE14: Revised code, sec1+2, entering sec3+4

--- 2014-03-13 sciunit-icse14/sciunit-icse14.tex
@@
 In this paper, we introduce \emph{SciUnit}, a framework for test-driven scientific model validation and outline how SciUnit, supported by new and existing collaborative infrastructure, could be  integrated into the modern scientific workflow.
-%in the context of the applications and architectures that the language may need to target.
-%
-%Programming languages targeted toward end-users in high-performance computing must consider a number of issues in addition to performance.
-Typically, researchers have focused verifiability, portability and ease-of-use, and the right balance often depends on the particular uses that they are put toward.
-%
-%Ace %, evaluates it against several design and adoption criteria, and describes several use cases as well as a realistic case study involving parallelizing neurobiological circuit simulations on clusters of GPUs.
-%is an extensible statically-typed programming language embedded within the ubiquitous scripting language Python.
-Python serves as a {metalanguage}, enabling programmatic control over several aspects of language specification and compilation.
-%We demonstrate the power of this design for HPC by defining all of the primitives of the OpenCL kernel language as a library of user-defined types in Ace, then extending this core with several higher-level first-class abstractions.
-The target of translation, called the \emph{backend}, is user-defined as well, allowing us to implement the semantics of these primitives by direct translation to OpenCL, CUDA or C99.
-%
-%Although Ace is built around a static type discipline, it supports a novel form of generic programming based on type propagation that, together with a form of extensible local type inference, eliminates most of the syntactic overhead usually associated with statically-typed languages.
-A companion host API allows type information from \verb|numpy| arrays to directly propagate into Ace kernels, supporting seamless adoption of Ace into current scientific workflows.
-%
 \end{abstract}
 \category{D.2.5}{Software Engineering}{Testing and Debugging}%[complexity measures, performance measures]
-\terms{Documentation, measurement, standardization}
-\keywords{unit testing, model validation, cyberinfrastructure}
+\terms{Documentation; measurement; standardization}
+\keywords{unit testing; model validation; cyberinfrastructure}
 
@@
 \includegraphics[scale=0.45]{diagram1.pdf}
-\vspace{-40px}
-\caption{\small Tests are derived from data and models are derived from scientific theories.
+\vspace{-50px}
+\caption{\small Tests are derived from data and models are derived from scientific theories, either directly or through existing data and model repositories.
 The score table summarizes the performance of a collection of models against a suite of tests.
-A table is generated and visualized in an IPython notebook stored inside a \emph{suite repository} (e.g.
-\texttt{saturnsuite}) hosted on social coding infrastructure (here, Github).
-SciDash is a portal that discovers and organizes these suite repositories and provides read-only views of the notebooks they contain.
-Common testing code, capabilities and bridges to external model and data repositories are also collaboratively developed in \emph{common repositories} (e.g.
+Tests and models are initialized, executed and visualized using \texttt{sciunit} in an IPython notebook stored inside a \emph{suite repository} (e.g.
+\texttt{saturnsuite}) hosted on a social coding website (e.g.
+Github).
+SciDash  discovers and organizes  suite repositories and provides interactive views of the notebooks they contain.
+Common testing code, capabilities and bridges to external model and data repositories are also collaboratively developed (e.g.
 \texttt{cosmounit}).}
@@
 Quantitative models are today validated by \emph{peer review}.
-For a model to be accepted by a scientific community, its advocates must submit a paper describing how it works and providing evidence that it predicts a quantity of interest more accurately than previous models, or in some cases, that it makes a desirable tradeoff  between accuracy and complexity \cite{box1987empirical}.
-Other members of the relevant community are then tasked with ensuring that validity was measured properly and that relevant data and competing models were adequately considered, drawing on knowledge of statistical methods and on the prior literature.
+For a model to be accepted by a scientific community, its advocates must submit a paper that describes how it works and provides evidence that it makes more accurate predictions than previous models (or that it makes a desirable tradeoff  between accuracy and complexity) \cite{box1987empirical}.
+Other members of the relevant community are then tasked with ensuring that validity was measured properly and that all relevant data and competing models were adequately considered, drawing on knowledge of statistical methods and the prior literature.
 Publishing is a primary motivator for most scientists \cite{howison2011scientific}.
 
-Quantitative scientific modeling and software development share much in common.
+Quantitative scientific modeling shares much in common with software development.
 Indeed, quantitative models are increasingly being implemented in software and in some cases, the software \emph{is} the model (e.g.
@@
 %A suite of such tests can be considered a partial specification of the program or component under development.  
-A suite of such tests complements code review by making it easier to answer questions like these more easily:
+A suite of such tests complements code review by making it easier to answer questions like these:
 \begin{enumerate}
@@
 \item What are the contemporary community standards for measuring goodness-of-fit?
-\item Which observations are already explained by existing models? What are the best models? What are the open modeling problems of interest?
+\item Which observations are already explained by existing models?
+What are the best models for a particular quantity?
+What are the open modeling problems of interest?
+What data has yet to be explained?
 %\item How much better is a new model than the state of the art? What are the open problems in my field?
-\item How do newly-made experimental observations impact the validity of previously-published models? Can new models explain previously published data?
+\item How do newly-made experimental observations impact the validity of previously published models? Can new models explain previously published data?
 \end{enumerate}
-But while software engineers can rely on a program's test suite, scientists today must extract this information from a body of scientific publications.
+But while software engineers can rely on a program's test suite, scientists today must extract this information from a body of scientific papers.
 This is increasingly difficult.
-Each publication focuses on just one model and is frozen in time, so it does not consider the latest experimental data or statistical methods.
+Each paper is narrowly focused, often considering just one model or dataset, and is frozen in time, so it does not consider the latest experimental data or statistical methods.
 Discovering, precisely characterizing and comparing models to discover the state of the art and find open modeling problems can require an encyclopedic knowledge of the literature, as can finding all data relevant to a model.
-Senior scientists often attempt to fill this need by publishing review papers, but in many areas, the number of publications generated every year can be overwhelming \cite{jinha_article_2010}, and comprehensive reviews of a particular area are published relatively infrequently.
+Senior scientists often attempt to fill this need by publishing review papers, but in many areas, the number of publications generated every year can be overwhelming \cite{jinha_article_2010}, and comprehensive reviews of a particular area are published infrequently.
 Statisticians often complain that scientists are not following best practices and that community standards evolve too slowly because a canonical paper or popular review used outdated methods.
@@
 
-One might compare this to a large software development team answering the questions listed above based only on carefully reviewed component documentation together with annual high-level project summaries.
-Although certainly a caricature, this motivates our suggestion that the scientific process could be improved by the adoption of test-driven methodologies alongside traditional peer review.
+One might compare this to a large software development team answering the questions listed above based largely on carefully reviewed, but rarely updated,  API documentation and annual high-level project summaries.
+While certainly a caricature, this motivates our suggestion that the scientific process could be improved by the adoption of test-driven methodologies alongside traditional peer review.
 However, the scientific community presents several unique challenges that must be addressed by any such methodology:
 \begin{enumerate}
-\item Unit tests are typically pass/fail, while goodness-of-fit between a model and data is typically measured by a continuous metric (e.g. a $p$-value). 
-\item Unit tests often test a \emph{particular} component, whereas a \emph{validation test} must be able to handle and compare many models capable of predicting the same quantity.
+\item Unit tests are typically pass/fail, while goodness-of-fit between a model and data is typically measured by a continuous metric. 
+\item Unit tests often test a \emph{particular} component, whereas a \emph{validation testing} system must be able to handle and compare many models with the same scope, developed by teams who only loosely coordinate their efforts.
 %They are sometimes reluctant to share certain details of their implementations \cite{badscientists}.
 %\item Experimental data can itself be erroneous, noisy or inconsistent with other available data.
-\item Different modelers use different programming languages.
+\item Scientists use a variety of programming languages.
 \item Professional software developers are typically trained in testing practices and tools, while scientists  rarely have training or experience with testing practices \cite{oai:open.ac.uk.OAI2:17673}.
-Thus, the framework must be as simple as possible.
-\item Different communities, groups and individuals prefer different goodness-of-fit metrics and care about different sets of observable quantities.
+%Thus, the framework must be as simple as possible.
+\item Different communities, groups and individuals prefer different goodness-of-fit metrics and focus on different sets of observable quantities.
 In contrast, there is typically more pressure to agree upon requirements and priorities in a software development project.
@@
 
-To address challenges 1-4, we will introduce a lightweight scientific validation testing framework, \textit{SciUnit}, in Sec.
+To begin to address these challenges, we will introduce a lightweight and flexible validation testing and framework called \textit{SciUnit} in Sec.
 2.
-Challenge 5 has to do with coordination between scientists.
-To address this, we introduce a community workflow based on widely-adopted social coding tools (here Github) along with a lightweight community portal called \textit{SciDash}.
+Many of these challenges have to do with coordination between scientists.
+To begin to address this, we then describe a community workflow based on widely-adopted social coding tools (here, Github) and a lightweight community portal called \textit{SciDash}, in Sec.
+\ref{scidash}.
 The overall goal of this work is to help scientists generate and examine tables like the one central to Figure 1, where the relative validity of a set of models having a common scope can be determined by examining scores produced by a suite of validation tests constructed from experimental data.
-
+We discuss adoption strategies and directions for future research into scientific model validation practices in Sec. \ref{discussion}.
 %Anecdotally, one of the greatest fears of new scientists is that a problem that they are working on has already been solved and published.
@@
 \section{Validation Testing with {SciUnit}}
-\begin{figure}[t]
-\small
-\begin{python}
-class PositionTest(sciunit.Test):
-    """Tests a planetary position model based on positions observed on day n given the positions in the n-1 previous days.
-    Metric: Standard p-value.
-    Parameters: 
-      obs_histories : list[list[position]]
-      obs_positions: list[position]"""
-	def __init__(self, obs_histories, obs_positions):
-		self.obs_histories = obs_histories
-		self.obs_positions = obs_positions
-
-	required_capabilities = [PredictsPlanetaryPosition]
-
-	def _judge(self, model):
-		predictions = []
-		for obs_history in self.obs_histories:
-			predictions.append(model.predict_next_pos(obs_history))
-		p = pooled_p_val(predictions, self.obs_positions)
-		return sciunit.PValue(p, related_data={
-			'obs_histories': obs_histories,
-			'obs_positions': obs_positions,
-			'predictions': predictions
-		})
-\end{python}
-\vspace{-5px}
-\caption{An example test class in \texttt{cosmounit}.}
-\label{fig:rate_test}
-\vspace{-15px}
-\end{figure}
 
@@
 Before explaining the details of this example, we point out that SciUnit is implemented in Python.
-Python is one of the most widely used languages in science today \cite{sanner1999python}.
-It supports calling into  many of the other popular languages, including R, MATLAB, C and Java, more cleanly than these languages support calling into Python (challenge 3).
-It is widely-recognized as being easy-to-read and its object system can be used to define abstract interfaces, which we leverage in support of challenge 2.
+Python is one of the most widely used languages in science today \cite{sanner1999python}, and is increasingly regarded as the \textit{de facto} language of open source scientific computing.
+It is easy to read and has a simple object system that we use to specify the interface between tests and models (challenge 2).
+It supports calling into  many other popular languages, including R, MATLAB, C and Java, often more cleanly than they support calling into Python (challenge 3).
+The IPython notebook is a powerful web-based ``read-eval-visualize loop'' that we use to support interactive table visualization, and it also permits using other languages on a per-cell basis \cite{ipynb}.
+Together, this makes writing wrappers around tests and models written in other languages relatively simple.
+We anticipate leveraging this infrastructure to make generating such wrappers nearly automatic as future work.
+\begin{figure}[t]
+\small
+%class PositionTest(sciunit.Test):
+%    """Tests a planetary position model based on positions observed on night n given the positions in the n-1 previous nights.
+%    Metric: Standard p-value.
+%    Parameters: 
+%      obs_histories : list[list[position]]
+%      obs_positions: list[position]"""
+%	def __init__(self, obs_histories, obs_positions):
+%		self.obs_histories = obs_histories
+%		self.obs_positions = obs_positions
+%
+%	required_capabilities = [PredictsPlanetaryPosition]
+%
+%	def _judge(self, model):
+%		predictions = []
+%		for obs_history in self.obs_histories:
+%			predictions.append(model.predict_next_pos(obs_history))
+%		p = pooled_p_val(predictions, self.obs_positions)
+%		return sciunit.PValue(p, related_data={
+%			'obs_histories': obs_histories,
+%			'obs_positions': obs_positions,
+%			'predictions': predictions
+%		})
+\begin{python}
+class PositionTest(sciunit.Test):
+  """Tests a planetary position model based on positions observed on night n given the positions in the n-1 previous nights.
+  Observation: {
+      'histories':list[list[Position]], 
+      'positions':list[Position]]
+  }"""
+  required_capabilities = [PredictsPlanetaryPosition]
+  def generate_prediction(self, model):
+    return [model.predict_next_pos(obs_history)
+      for obs_history 
+      in self.observation['histories']]
 
-A \textit{SciUnit} validation test is an {instance} of a Python class implementing the \verbx{sciunit.Test} abstract interface (line 1). 
-Here, the class \verbx{PositionTest} takes two \emph{parameters} in its constructor (constructors are named \verbx{\_\_init\_\_} in Python, lines 7-9). 
-The meaning of each parameter along with a description of the goodness-of-fit metric used by the test is documented on lines 2-6. 
+  def compute_score(self, observation, prediction):
+    obs_positions = observation['positions']
+    return pooled_p_val([abs_distance(obs, pred)
+      for (obs, pred) 
+      in zip(obs_positions, prediction)])
+\end{python}
+\vspace{-12px}
+\caption{\small An example test class in \texttt{cosmounit}.}
+\label{fig:rate_test}
+\vspace{-15px}
+\end{figure}
+
+A \textit{SciUnit} validation test is an {instance} of a Python class implementing the \verbx{sciunit.Test} interface (line 1). 
+A test takes one required parameter in its constructor (inherited from the base class) containing the observations against which the test will validate models.
+For the class \verbx{PositionTest}, this observation parameter is a dictionary containing two types of observations (documented per Python conventions on lines 2-6). 
 %For convenience, we also make use of functions provided by the popular NumPy\cite{numpy_url} and SciPy\cite{scipy_url} libraries, although these are not required by \textit{SciUnit}.  
-To create a \emph{particular} position test, we instantiate this class with particular planetary observations. 
-For example, the subset of cosmologists interested specifically in Saturn might instantiate a test by randomly chunking observations made about Saturn as follows:
+To create a \emph{particular} position test, we instantiate this class with particular observations. 
+For example, the subset of cosmologists interested specifically in Saturn might instantiate a test by randomly chunking observations made about Saturn's nightly position over time as follows:
+\vspace{-3px}
 \begin{python}
   h, p = randomly_chunk(saturn_obs_positions)
-  saturn_position_test = PositionTest(h, p)
+  saturn_position_test = PositionTest(
+    {'histories': h, 'positions': p})
+\end{python}
+\vspace{-5px}
+The class \verb|PositionTest| defines logic that is not specific to any particular planet, so it is contained in a repository shared by all cosmologists called \verbx{cosmounit}.
+The particular test above would be constructed in a separate repository focused specifically on Saturn called \verbx{saturnsuite}.
+Both would be collaboratively developed by these (overlapping) research communities in source code repositories on Github.
+
+Classes that implement the \verbx{sciunit.Test} interface must contain a \verbx{generate_prediction} method that takes a \emph{model} as input and extracts a prediction.
+To specify the interface between the test and the model, the test author specifies a list of \emph{required capabilities} in the \verbx{required_capabilities}, seen on line 7.
+A capability specifies a collection of methods that a test will need to invoke in order to receive relevant data, and is analogous to an \emph{interface} in a language like Java. 
+In Python, interfaces are written as classes with unimplemented members. 
+The capability required in Figure \ref{fig:rate_test} is shown in Figure \ref{fig:capability}. 
+Classes defining capabilities are tagged as such by inheriting from \verbx{sciunit.Capability}.
+The test in Figure \ref{fig:rate_test} repeatedly uses this capability on lines 9-11 to return a list of predicted positions for each observed history (using a Python list comprehension for brevity).
+A planet's position is represented using a standard celestial coordinate system specified within \verbx{cosmounit} by the class \verbx{Position}.
+A model class can implement a capability by simply inheriting from it and implementing the required methods (Figure \ref{fig:capability}).
+A particular model is an instance of such a class.
+The scope of a model is identified by the capabilities it implements.
+\vspace{-3px}
+\begin{python}
+lin_saturn_model = LinearPlanetModel()
+\end{python}
+\vspace{-5px}
+Once a prediction has been extracted from a model, the test class must compute a \emph{score}.
+The framework invokes the \verbx{compute_score} method with the observation provided upon test instantiation and the prediction just generated (line 14 of Fig.
+\ref{fig:rate_test}).
+Here, the test class constructs a list of distances between the observed and predicted positions, then determines a pooled $p$-value to determine the goodness-of-fit (the details are omitted for concision).
+A $p$-value is represented as an instance of \verbx{sciunit.PValue}, a subclass of \texttt{sciunit.Score} that has been included with {SciUnit} due to its wide use across science.
+%In addition to the $p$-value itself, the returned score object also contains metadata, accessible via the \verbx{related_data} attribute, for  scientists who may wish to examine the result in more detail later.
+This includes the observation, prediction, and a reference to the model and test.
+\begin{figure}
+\small
+\begin{python}
+class PredictsPlanetaryPosition(sciunit.Capability):
+  def predict_next_pos(self, history): 
+    """Takes a list of previous Positions and produces the next Position."""
+    raise NotImplementedError()
+
+class LinearPlanetModel(sciunit.Model, 
+    PredictsPlanetaryPosition):
+  def predict_next_pos(self, history):
+    return linear_prediction(history)
+\end{python}
+\vspace{-12px}
+\caption{\small An example capability and a model class that implements it in \verbx{cosmounit}.}
+\label{fig:capability}
+\vspace{-10px}
+\end{figure}
+
+This illustrates a key difference between unit testing, which would simply produce a boolean result, and our conception of scientific validation testing (challenge 1).
+A score class must induce an ordering, so that a table like that shown in Figure 1 can be sorted along its columns, and it can  specify a normalization scheme so the cells can be color-coded.
+Predictions are extracted by a separate method to make it easier for statisticians to write new tests for which only the goodness-of-fit metric differs, not the method by which the prediction is extracted (challenge 5).
+%It also makes it easier to cache predictions generated by large-scale models so that they can be used by multiple tests.
+
+Te \verbx{judge} method of a test can be invoked to compute a score for a single model: 
+\vspace{-4px}
+\begin{python}
+score = saturn_position_test.judge(lin_saturn_model)
+\end{python}
+\vspace{-5px}
+This method proceeds by first checking that the provided model implements all required capabilities before calling the \verbx{generate_prediction} method followed immediately by the \verbx{compute_score} method.
+A reference to the test,   model, observation, prediction and other related data the test provides (not shown) are available as attributes of the score.
+
+To produce a more comprehensive test suite, the contributors to \verbx{saturnsuite} would instantiate a number of other tests and then create an instance of the \verbx{TestSuite} class:
+\begin{python}
+saturn_motion_suite = sciunit.TestSuite([
+  saturn_position_test, saturn_velocity_test, ...])
+\end{python}
+They would also instantiate a number of models.
+A test suite can be used to judge multiple models at once (challenge 2) if they satisfy the union of the capabilities required by the constituent tests.
+The result is a \emph{score matrix}:
+\begin{python}
+sm_matrix = saturn_motion.judge([lin_copernicus_sat, ptolemy_sat, newton_sat, kepler_sat])
 \end{python}
 
-The class \verb|PositionTest| defines logic that is not specific to any particular planet, so it is contained in a package shared by all cosmologists called \verbx{cosmounit}, while the particular test above would be used in a test suite focused specifically on Saturn called \verbx{saturnsuite}.
-Both the common logic and specific suites are collaboratively developed by these overlapping research communities in source code repositories on Github (or another similar service).
+If constructed inside an iPython notebook, a score matrix can then be visualized as an interactive table, much like the one shown in Figure \ref{fig:sciunit_overview}.
+Scientists can sort by column and click on tests, models and scores to get more information on each (not shown, implementation in progress).
 
-Classes that implement the \verbx{sciunit.Test} interface must contain a \verbx{_judge} method that receives a candidate \emph{model} as input and produces a \textit{score} as output. 
-To specify the interface between the test and the model, the test author provides a list of \emph{capabilities} in the \verbx{required_capabilities} attribute, seen on line 11 of Fig.
-\ref{fig:rate_test}.
-Capabilities are simply collections of methods that a test will need to invoke in order to receive relevant data, and are analogous to \emph{interfaces} in e.g. Java. 
-In Python, capabilities must be written as classes with unimplemented members. 
-The capability required by the test in Figure \ref{fig:rate_test} is shown in Figure \ref{fig:capability}. 
-In \textit{SciUnit}, classes defining capabilities are tagged as such by inheriting from \verbx{sciunit.Capability}.
-The test in Figure \ref{fig:rate_test} repeatedly uses this capability on line 16 to produce a position prediction for each observation.
-We assume that positions are represented in a standardized manner specified within \verbx{cosmounit}.
-A model is simply an object that implements capabilities (via Python's simple inheritance mechanism).
-
-The remainder of the \verbx{_judge} method compares the model predictions to observed data to produce a pooled $p$ value.
-The method returns an instance of \verbx{sciunit.PValue}, a subclass of \texttt{sciunit.Score} that has been included with \textit{SciUnit} due to its generality.
-In addition to the $p$-value itself, the returned score object also contains metadata, via the \verbx{related_data} parameter, for  scientists who may wish to examine the result in more detail later.
-This illustrates some key differences between unit testing, which would simply produce a boolean result, and our conception of scientific validation testing (challenge 1).
-A score must induce an ordering, so that the table shown in Figure 1 can be sorted along its columns, and it can optionally specify a normalization scheme so the cells can be color-coded (not shown).
-A test can be manually executed using the \verbx{judge} method:
-\begin{python}
-score = saturn_position_test.judge(kepler_sat)
-\end{python}
-
-This method proceeds by first checking that the provided model implements all required capabilities before calling the test's \verbx{_judge} method to produce a score.
-A reference to the test and model are added to the score for convenience (accessible via the \verbx{test} and \verbx{model} attributes, respectively).
-
-Alongside this position test, we could also provide a number of other test classes in \verbx{cosmounit} and instantiate them with data about aspects of Saturn's motion (e.g.
-it's velocity)  to produce a comprehensive suite in \verbx{saturnsuite}.
-\begin{python}
-saturn_motion = sciunit.TestSuite([saturn_position_test, saturn_velocity_test, ...])
-\end{python}
-Like a single test, a test suite is capable of judging one or more models. The result is a score matrix much like the one diagramed in Fig. \ref{fig:sciunit_overview}.
-\begin{python}
-sm_matrix = saturn_motion.judge([copernicus_sat, ptolemy_sat, newton_sat, kepler_sat])
-\end{python}
-A test suite requires the union of the capabilities required by the tests it contains.
-A model that performs well across tests in such a suite could reasonably claim (e.g.
-to reviewers) that it is a coherent, valid model of Saturn's motion.
-As new data is collected, new tests can be added to the suite.
-However, because the interface between the test and the model remains the same (only the data parameterizing the tests changes), the table can be updated automatically.
-Similarly, when a new model is developed, it can immediately be evaluated against all known data that has been encoded as a test by simply exposing its predictions via the capabilities the test requires.
 %They then defined a combined metric favoring models that broadly succeeded at meeting these criteria, to produce an overall ranking. 
@@
 %We discuss visualization of results in Secs. \ref{sec:scidash_activities} and \ref{sec:scidash_visualization}.
-
-\begin{figure}
-\begin{python}
-class PredictsPlanetaryPosition(sciunit.Capability):
-  def predict_next_pos(self, history): 
-    """Takes a list of previous positions and produces the next position."""
-    raise NotImplementedError("Model does not implement capability.")
-\end{python}
-\caption{An example capability specifying a single required method (used by the test in Figure \ref{fig:rate_test}).}
-\label{fig:capability}
-\vspace{-10px}
-\end{figure}
 
@@
 %
-\section{Collaborative Workflow}
+\section{Collaborative Workflow}\label{scidash}
+A model that performs well across tests in such a suite could reasonably claim (e.g.
+to reviewers) to be a coherent, valid model of Saturn's motion.
+As new data is collected, new tests can be added to the suite or existing tests can be refined.
+As long as the interface between the test and the model remains the same, the table can be updated entirely automatically.
+Similarly, when a new model is developed, it can immediately be evaluated against all known data that has been encoded as a test by simply exposing its predictions via the capabilities the test requires.
+If better statistical tests are developed, they too can be used to regenerate the table automatically, as discussed above.
+
 The design just described has been purposefully left as simple as possible, in pursuit of challenge 4.
@@
 
-\section{Discussion}
+\section{Discussion}\label{discussion}
 Academic peer review and code review are similar, but software developers typically supplement human review with test-driven methodologies.
@@
 3.
+
+We also plan to use SciUnit to implement competitions among classes of models or techniques in particular subfields.
+Typically such competitions use \emph{ad hoc} infrastructure.
+A common framework could be used to make designing and continuously operating such contests more feasible.
  %In particular, we consider modeling competitions in science as a form of test-driven validation.
@@
 \section*{Acknowledgements}
-We thank Sharon Crook, Shreejoy Tripathy and Padraig Gleeson for their support of this project.  
+We thank Sharon Crook, Shreejoy Tripathy and Padraig Gleeson for their helpful discussions.  
 The work was supported in part by grant R01MH081905 from the National Institute of Mental Health. 
@@
 % (used to reserve space for the reference number labels box)
-\bibliographystyle{IEEEtran}
+\bibliographystyle{abbrv}
 \bibliography{../research,references,references_crook,urls,references_ja}

### 2014-03-14 sciunit-icse14/sciunit-icse14.tex
ICSE14: Revised through most of sec3, added some TODOs

@@
 
-A \textit{SciUnit} validation test is an {instance} of a Python class implementing the \verbx{sciunit.Test} interface (line 1). 
-A test takes one required parameter in its constructor (inherited from the base class) containing the observations against which the test will validate models.
-For the class \verbx{PositionTest}, this observation parameter is a dictionary containing two types of observations (documented per Python conventions on lines 2-6). 
+A {SciUnit} validation test is an {instance} of a Python class implementing the \verbx{sciunit.Test} interface (line 1). 
+A test takes one required parameter in its constructor (inherited from the base class) containing the observation(s) against which the test will validate models.
+In our example, this is a dictionary documented per Python conventions on lines 2-6. 
 %For convenience, we also make use of functions provided by the popular NumPy\cite{numpy_url} and SciPy\cite{scipy_url} libraries, although these are not required by \textit{SciUnit}.  
@@
 They would also instantiate a number of models.
-A test suite can be used to judge multiple models at once (challenge 2) if they satisfy the union of the capabilities required by the constituent tests.
+A test suite can be used to judge multiple models at once (challenge 2), if they satisfy the union of the capabilities required by the constituent tests.
 The result is a \emph{score matrix}:
 \begin{python}
-sm_matrix = saturn_motion.judge([lin_copernicus_sat, ptolemy_sat, newton_sat, kepler_sat])
+sm_matrix = saturn_motion_suite.judge([lin_saturn_model, ptolemy_model, kepler_model])
 \end{python}
 
-If constructed inside an iPython notebook, a score matrix can then be visualized as an interactive table, much like the one shown in Figure \ref{fig:sciunit_overview}.
-Scientists can sort by column and click on tests, models and scores to get more information on each (not shown, implementation in progress).
+If constructed inside an IPython notebook, a score matrix can then be visualized as an interactive table, much like the one shown in Figure \ref{fig:sciunit_overview}.
+Scientists can sort by column and click on tests, models and scores to get more information on each, including documentation, authorship information and related data extracted from the underlying objects.
 
@@
 \section{Collaborative Workflow}\label{scidash}
-A model that performs well across tests in such a suite could reasonably claim (e.g.
-to reviewers) to be a coherent, valid model of Saturn's motion.
-As new data is collected, new tests can be added to the suite or existing tests can be refined.
-As long as the interface between the test and the model remains the same, the table can be updated entirely automatically.
-Similarly, when a new model is developed, it can immediately be evaluated against all known data that has been encoded as a test by simply exposing its predictions via the capabilities the test requires.
-If better statistical tests are developed, they too can be used to regenerate the table automatically, as discussed above.
 
-The design just described has been purposefully left as simple as possible, in pursuit of challenge 4.
-Despite its simple design, it captures the essential elements of the scientific model validation process and satisfies the four challenges we laid out.
-SciUnit can be used by individual scientists to organize their workflows, but because science, and in particular, model validation is a collaborative process, we have also described the intended use of SciUnit within a collaborative workflow, mediated by a social coding tool like Github.
+Social coding tools like Github allow each scientist to work on their own fork of a repository where they can examine the consequences of their own contributions on  the tests and models that have previously been developed, facilitating both model and data sharing, and the organic adoption of community standards.
+For example, an experimentalist who has gathered new data can instantiate a new test or refine an existing one  with more accurate observations.
+As long as the interface between the test and the model remains the same, the scores can be recomputed entirely automatically, requiring no coordination with the original model developers (challenge 2).
+Similarly, when a modeler develops a new model, she can immediately  validate it against all the tests it is capable of taking.
+If a statistician develops a better goodness-of-fit metric, it too can be turned into a collection of tests that uses existing capabilities, and so the table can be regenerated automatically, as discussed above.
+If a scientist is not interested in certain tests, she can locally remove them to recalibrate her view of the relative merits of different models (challenge 5).
 
-More directly: we anticipate common testing logic (e.g.
+Once satisfied, these scientists might submit a \emph{pull request} to the central repository (either separately or in conjunction with the submission of a conventional paper).
+Anonymous reviewers tasked by the ``editors'' of the repository would then subject the pull request to appropriate peer review before accepting it into the ``academic record''.
+A model that performs well across tests in a collaboratively developed suite could more easily claim (e.g.
+to reviewers) to be a coherent, valid model of, for example, Saturn's motion.
+
+%Because new models, tests and data are being added by a variety of people, there needs to be a collaborative workflow in place.
+As we have outlined above, the basic mechanisms have already been put in place by software developers.
+
+To help help scientists discover these repositories and agree on which ones are central, we are developing a collaborative portal called \textit{SciDash} (\url{http://scidash.org/}).
+It is essentially a wiki pointing scientists to the $X$\verbx{unit}and $X$\verbx{suite} repositories designated as ``primary'' by early adopters in their research areas (who serve as \emph{de facto} initial editors as well).
+SciDash also supports loading IPython notebooks containing generated tables stored in these repositories into the browser directly so that score tables can be explored without going to the trouble of cloning the repository locally.
+To modify the suite (e.g.
+by adding the model a scientist is developing to it), a scientist can simply fork the repository by clicking a button at the top of such a view.
+We are exploring ways of loading forks directly into a cloud-based IPython notebook system, so that the complexities of Git and installation of Python and the packages required by the repository are left behind the scenes (challenge 4).
+SciDash discovers public forks of repositories that it is indexing and lists them to help scientists get a better sense of who is working on the problems they are interested in and what partial results are publicly available.
+
+TODO: How to get adoption?
+Run contests -- dash double meaning.
+Also ask people directly (worked for Scholarpedia...) We also plan to use SciUnit to implement competitions among classes of models or techniques in particular subfields.
+Typically such competitions use \emph{ad hoc} infrastructure.
+A common framework could be used to make designing and continuously operating such contests more feasible.
+
+%More directly: we anticipate common testing logic (e.g.
 \verbx{PositionTest}), modeling logic and capabilities being collaboratively developed by larger communities in less specialized repositories like \verbx{cosmounit}.
@@
 % This workflow is meant to evoke the \emph{test-driven development} methodology that is often used in software projects.
-
-Statisticians wishing to promote new validation metrics can simply for a $X$\verbx{unit} repository and implement new test logic.
+%
+%Statisticians wishing to promote new validation metrics can simply for a $X$\verbx{unit} repository and implement new test logic.
 By reusing the same interfaces as previous tests used, these new metrics can immediately be used to validate or invalidate a large body of existing models, providing valuable data for use in convincing the community to adopt these into the mainstream repositories.
@@
 
-To help organize these repositories, a simple collaborative portal sitting above Github called \textit{SciDash} is currently under development (\url{http://scidash.org/}).
-SciDash consists of an organized, collaboratively filtered listing of $X$\verbx{unit} and $X$\verbx{suite} repositories.
-That is, it is a tool to help scientists find the most popular repositories in their research area, to facilitate the development of community standards.
-SciDash also supports extracting summaries of tests, models, scores and related data from suite repositories to generate hyperlinked score tables and documentation.
-This facilitates exploratory workflows.
-To modify a suite (e.g.
-by adding the model a scientist is developing to it), a scientist can simply fork the repository.
-SciDash automatically discovers public forks of repositories that it is already indexing.
+\section{Discussion}\label{discussion}
+The design just described has been purposefully left as simple as possible, in pursuit of challenge 4.
+Despite its simple design, it captures the essential elements of the scientific model validation process and satisfies the four challenges we laid out.
+TODO: SciUnit is essentially an embedded DSL that captures precisely the standard workflow of model validation.
+SciUnit can be used by individual scientists to organize their workflows, but because science, and in particular, model validation is a collaborative process, we have also described the intended use of SciUnit within a collaborative workflow, mediated by a social coding tool like Github.
 
-\section{Discussion}\label{discussion}
-Academic peer review and code review are similar, but software developers typically supplement human review with test-driven methodologies.
+%Academic peer review and code review are similar, but software developers typically supplement human review with test-driven methodologies.
 Such methodologies may benefit scientific communities as well, but several unique  constraints make reusing existing testing frameworks and processes difficult.
@@
 We have used this as part of a collaboration with a large-scale neuroinformatics project (http://www.opensourcebrain.org), whose developers are enthusiastic about a practical pipeline for testing the dozens of models they currently maintain.
-Much future work remains to be done to investigate whether these tools are truly usable and useful to scientific communities, to further develop the community workflow and infrastructure and to develop a range of realistic case studies.
+TODO: More SE-research-friendly future work discussion Much future work remains to be done to investigate whether these tools are truly usable and useful to scientific communities, to further develop the community workflow and infrastructure and to develop a range of realistic case studies.
 Nevertheless, we believe that identifying the synergies between testing practices and model validation, and describing the basic tooling, represents a novel contribution to the literature on software testing.
@@
 
-We also plan to use SciUnit to implement competitions among classes of models or techniques in particular subfields.
-Typically such competitions use \emph{ad hoc} infrastructure.
-A common framework could be used to make designing and continuously operating such contests more feasible.
  %In particular, we consider modeling competitions in science as a form of test-driven validation.
@@
 We thank Sharon Crook, Shreejoy Tripathy and Padraig Gleeson for their helpful discussions.  
-The work was supported in part by grant R01MH081905 from the National Institute of Mental Health. 
-The content is solely the responsibility of the authors and does not necessarily represent the official views of the National Institutes of Health.
+The work was supported in part by NIMH grant R01MH081905. 
+The content is solely the responsibility of the authors and does not necessarily represent the official views of the NIH.
 

### 2014-03-14 sciunit-icse14/sciunit-icse14.tex
ICSE14: revised sec1+2

@@
 \begin{enumerate}
-\item What functionality should each component have?
+\item What functionality is a component expected to have?
 \item What functionality has been adequately implemented? What remains to be done?
@@
 \begin{enumerate}
-\item What are the contemporary community standards for measuring goodness-of-fit?
-\item Which observations are already explained by existing models?
-What are the best models for a particular quantity?
-What are the open modeling problems of interest?
-What data has yet to be explained?
+\item What is a model's scope and how is validity measured?
+\item Which observations are already explained by existing models? What are the best models of a particular quantity?  What data has yet to be explained?
 %\item How much better is a new model than the state of the art? What are the open problems in my field?
-\item How do newly-made experimental observations impact the validity of previously published models? Can new models explain previously published data?
+\item What affect do new observations have on the validity of previously published models? Can new models explain previously published data?
 \end{enumerate}
@@
 This is increasingly difficult.
-Each paper is narrowly focused, often considering just one model or dataset, and is frozen in time, so it does not consider the latest experimental data or statistical methods.
+Each paper is narrowly focused, often considering just one model or dataset, and is frozen in time, so it does not consider the latest models, data or statistical methods.
 Discovering, precisely characterizing and comparing models to discover the state of the art and find open modeling problems can require an encyclopedic knowledge of the literature, as can finding all data relevant to a model.
@@
 Statisticians often complain that scientists are not following best practices and that community standards evolve too slowly because a canonical paper or popular review used outdated methods.
-Furthermore, if the literature simply doesn't address an important question of validity, because  new data has been gathered since the model paper was published for example, a researcher might need to reimplement an existing model in order to evaluate it.
+Furthermore, if the literature simply doesn't address an important question of validity, a researcher might need to reimplement an existing model in order to evaluate it.
 
@@
 While certainly a caricature, this motivates our suggestion that the scientific process could be improved by the adoption of test-driven methodologies alongside traditional peer review.
-However, the scientific community presents several unique challenges that must be addressed by any such methodology:
+However, the scientific community presents several unique challenges:
 \begin{enumerate}
 \item Unit tests are typically pass/fail, while goodness-of-fit between a model and data is typically measured by a continuous metric. 
-\item Unit tests often test a \emph{particular} component, whereas a \emph{validation testing} system must be able to handle and compare many models with the same scope, developed by teams who only loosely coordinate their efforts.
+\item Unit tests often test a \emph{particular} component, whereas a \emph{validation testing} system must be able to handle and compare many models with the same scope.
 %They are sometimes reluctant to share certain details of their implementations \cite{badscientists}.
@@
 \item Scientists use a variety of programming languages.
-\item Professional software developers are typically trained in testing practices and tools, while scientists  rarely have training or experience with testing practices \cite{oai:open.ac.uk.OAI2:17673}.
+\item Scientists often only loosely coordinate their efforts.
+\item Different communities, groups and individuals prefer different goodness-of-fit metrics and focus on different sets of observable quantities.
+In contrast, there is more pressure to agree upon requirements and priorities in a typical software development project.
+\item Professional software developers are typically trained in testing practices and the use of collaborative tools, while scientists more rarely have such experience \cite{oai:open.ac.uk.OAI2:17673}.
 %Thus, the framework must be as simple as possible.
-\item Different communities, groups and individuals prefer different goodness-of-fit metrics and focus on different sets of observable quantities.
-In contrast, there is typically more pressure to agree upon requirements and priorities in a software development project.
 \end{enumerate}
@@
 Python is one of the most widely used languages in science today \cite{sanner1999python}, and is increasingly regarded as the \textit{de facto} language of open source scientific computing.
-It is easy to read and has a simple object system that we use to specify the interface between tests and models (challenge 2).
-It supports calling into  many other popular languages, including R, MATLAB, C and Java, often more cleanly than they support calling into Python (challenge 3).
-The IPython notebook is a powerful web-based ``read-eval-visualize loop'' that we use to support interactive table visualization, and it also permits using other languages on a per-cell basis \cite{ipynb}.
+It is easy to read and has a simple object system that we use to specify the interface between tests and models (challenge 2, see below).
+It supports calling into other popular languages, including R, MATLAB, C and Java, often more cleanly than they support calling into Python (challenge 3).
+The IPython notebook is a powerful web-based ``read-eval-visualize loop'' that we will use to support interactive table visualization, and it also permits using other languages on a per-cell basis \cite{ipynb}.
 Together, this makes writing wrappers around tests and models written in other languages relatively simple.
-We anticipate leveraging this infrastructure to make generating such wrappers nearly automatic as future work.
+We anticipate making it nearly automatic as future work.
 \begin{figure}[t]
@@
 
-A {SciUnit} validation test is an {instance} of a Python class implementing the \verbx{sciunit.Test} interface (line 1). 
-A test takes one required parameter in its constructor (inherited from the base class) containing the observation(s) against which the test will validate models.
-In our example, this is a dictionary documented per Python conventions on lines 2-6. 
+A {SciUnit} validation test is an {instance} of a Python class inheriting from  \verbx{sciunit.Test} (line 1). 
+Every test takes one required constructor argument (inherited from the base class): the observation(s) against which the test will validate models.
+In our example, this is a dictionary, documented per Python conventions on lines 2-6. 
 %For convenience, we also make use of functions provided by the popular NumPy\cite{numpy_url} and SciPy\cite{scipy_url} libraries, although these are not required by \textit{SciUnit}.  
@@
 \begin{python}
-  h, p = randomly_chunk(saturn_obs_positions)
+  h, p = randomly_chunk(obsvd_saturn_positions)
   saturn_position_test = PositionTest(
@@
 
-Classes that implement the \verbx{sciunit.Test} interface must contain a \verbx{generate_prediction} method that takes a \emph{model} as input and extracts a prediction.
-To specify the interface between the test and the model, the test author specifies a list of \emph{required capabilities} in the \verbx{required_capabilities}, seen on line 7.
-A capability specifies a collection of methods that a test will need to invoke in order to receive relevant data, and is analogous to an \emph{interface} in a language like Java. 
-In Python, interfaces are written as classes with unimplemented members. 
-The capability required in Figure \ref{fig:rate_test} is shown in Figure \ref{fig:capability}. 
+Classes that implement the \verbx{sciunit.Test} interface must specify two methods: a method for extracting a prediction from a model, and a method for producing a score from that prediction.
+Predictions are extracted by a separate method to make it easier for statisticians to write new tests for which only the goodness-of-fit metric differs, not the method by which the prediction is extracted (challenge 5).
+
+The \verbx{generate_prediction} method (lines 8-11) is passed a \emph{model} as input and must extract a \emph{prediction}.
+A model is an instance of a class inheriting from \verbx{sciunit.Model} and a prediction can be any Python data structure.
+To specify the interface between the test and the model, the test class specifies a list of \verbx{required_capabilities} (line 7).
+A capability specifies the methods that a test may need to invoke on the model to extract a prediction, and is analogous to an \emph{interface} in a language like Java. 
+In Python, capabilities are written as classes with unimplemented members, shown in Figure \ref{fig:capability}. 
 Classes defining capabilities are tagged as such by inheriting from \verbx{sciunit.Capability}.
-The test in Figure \ref{fig:rate_test} repeatedly uses this capability on lines 9-11 to return a list of predicted positions for each observed history (using a Python list comprehension for brevity).
+The test in Figure \ref{fig:rate_test} repeatedly uses this capability on lines 9-11 to return a list of predicted positions for each observed history (using a list comprehension for brevity).
 A planet's position is represented using a standard celestial coordinate system specified within \verbx{cosmounit} by the class \verbx{Position}.
-A model class can implement a capability by simply inheriting from it and implementing the required methods (Figure \ref{fig:capability}).
-A particular model is an instance of such a class.
-The scope of a model is identified by the capabilities it implements.
+
+A model class implements a capability by inheriting from it and implementing the required methods (Figure \ref{fig:capability}).
+The scope of a model class is identified by the capabilities it has.
+A particular model is an instance of such a class:
 \vspace{-3px}
@@
 \vspace{-5px}
+
 Once a prediction has been extracted from a model, the test class must compute a \emph{score}.
-The framework invokes the \verbx{compute_score} method with the observation provided upon test instantiation and the prediction just generated (line 14 of Fig.
-\ref{fig:rate_test}).
-Here, the test class constructs a list of distances between the observed and predicted positions, then determines a pooled $p$-value to determine the goodness-of-fit (the details are omitted for concision).
+The framework invokes the \verbx{compute_score} method with the observation provided upon test instantiation and the prediction just generated.
+On lines 13-17 of Fig.
+\ref{fig:rate_test}, the test class constructs a list of distances between the observed and predicted positions, then determines a pooled $p$-value to determine the goodness-of-fit (the details are omitted for concision).
 A $p$-value is represented as an instance of \verbx{sciunit.PValue}, a subclass of \texttt{sciunit.Score} that has been included with {SciUnit} due to its wide use across science.
@@
 A score class must induce an ordering, so that a table like that shown in Figure 1 can be sorted along its columns, and it can  specify a normalization scheme so the cells can be color-coded.
-Predictions are extracted by a separate method to make it easier for statisticians to write new tests for which only the goodness-of-fit metric differs, not the method by which the prediction is extracted (challenge 5).
 %It also makes it easier to cache predictions generated by large-scale models so that they can be used by multiple tests.
@@
 This method proceeds by first checking that the provided model implements all required capabilities before calling the \verbx{generate_prediction} method followed immediately by the \verbx{compute_score} method.
-A reference to the test,   model, observation, prediction and other related data the test provides (not shown) are available as attributes of the score.
+A reference to the test,   model, observation, prediction and other related data the test provides (none here) are available as attributes of the score.
 
-To produce a more comprehensive test suite, the contributors to \verbx{saturnsuite} would instantiate a number of other tests and then create an instance of the \verbx{TestSuite} class:
+To produce a comprehensive test suite, the contributors to \verbx{saturnsuite} would instantiate a number of other tests and then create an instance of the \verbx{TestSuite} class:
 \begin{python}
@@
 They would also instantiate a number of models.
-A test suite can be used to judge multiple models at once (challenge 2), if they satisfy the union of the capabilities required by the constituent tests.
+A test suite can be used to judge multiple models at once, if they satisfy the union of the capabilities required by the constituent tests (challenge 2).
 The result is a \emph{score matrix}:
 \begin{python}
-sm_matrix = saturn_motion_suite.judge([lin_saturn_model, ptolemy_model, kepler_model])
+sm_matrix = saturn_motion_suite.judge([
+  lin_saturn_model, ptolemy_model, kepler_model, ...])
 \end{python}
@@
 If constructed inside an IPython notebook, a score matrix can then be visualized as an interactive table, much like the one shown in Figure \ref{fig:sciunit_overview}.
-Scientists can sort by column and click on tests, models and scores to get more information on each, including documentation, authorship information and related data extracted from the underlying objects.
+Scientists can sort by column and click on tests, models and scores to get more information on each, including documentation, authorship information, related papers and other related data extracted from the underlying classes (details omitted for concision).
 
@@
 Similarly, when a modeler develops a new model, she can immediately  validate it against all the tests it is capable of taking.
-If a statistician develops a better goodness-of-fit metric, it too can be turned into a collection of tests that uses existing capabilities, and so the table can be regenerated automatically, as discussed above.
+If a statistician develops a better goodness-of-fit metric, it too can be turned into a collection of tests that use existing capabilities, and so the table can be regenerated automatically, as discussed above.
 If a scientist is not interested in certain tests, she can locally remove them to recalibrate her view of the relative merits of different models (challenge 5).
@@
 Once satisfied, these scientists might submit a \emph{pull request} to the central repository (either separately or in conjunction with the submission of a conventional paper).
-Anonymous reviewers tasked by the ``editors'' of the repository would then subject the pull request to appropriate peer review before accepting it into the ``academic record''.
+Anonymous reviewers tasked by the ``editors'' of the repository would then subject the pull request to appropriate peer review before committing it to the ``academic record''.
 A model that performs well across tests in a collaboratively developed suite could more easily claim (e.g.

### 2014-03-14 sciunit-icse14/sciunit-icse14.tex
ICSE14: Final draft

@@
 \usepackage[compact]{titlesec}
-\titlespacing{\section}{0pt}{*0}{*0}
+\titlespacing{\section}{0pt}{*0.2}{*0}
+%\titleformat{\section}
+%  {\large\bfseries\uppercase}{\thesection}{1em}{}
 \titlespacing{\subsection}{0pt}{*0}{*0}
@@
 \category{D.2.5}{Software Engineering}{Testing and Debugging}%[complexity measures, performance measures]
-\terms{Documentation; measurement; standardization}
-\keywords{unit testing; model validation; cyberinfrastructure}
+\terms{Documentation, Measurement, Standardization}
+\keywords{unit testing, model validation, cyberinfrastructure}
 
@@
 %\IEEEpeerreviewmaketitle
-
+\titleformat{\section}
+ {\secfnt}{\thesection}{1em}{\secfnt\MakeUppercase}
+ \vspace{10pt}
 \section{Introduction}
@@
 It supports calling into other popular languages, including R, MATLAB, C and Java, often more cleanly than they support calling into Python (challenge 3).
-The IPython notebook is a powerful web-based ``read-eval-visualize loop'' that we will use to support interactive table visualization, and it also permits using other languages on a per-cell basis \cite{ipynb}.
+The IPython notebook is a powerful web-based ``read-eval-visualize loop'' that we will use to support interactive table visualization, and it also permits using other languages on a per-cell basis \cite{perez2007ipython}.
 Together, this makes writing wrappers around tests and models written in other languages relatively simple.
@@
 
-Social coding tools like Github allow each scientist to work on their own fork of a repository where they can examine the consequences of their own contributions on  the tests and models that have previously been developed, facilitating both model and data sharing, and the organic adoption of community standards.
+Social coding tools like Github allow each scientist to work on their own fork of a repository like \verbx{saturnunit} where they can examine the consequences of their own contributions on  the tests and models that have previously been developed.
 For example, an experimentalist who has gathered new data can instantiate a new test or refine an existing one  with more accurate observations.
-As long as the interface between the test and the model remains the same, the scores can be recomputed entirely automatically, requiring no coordination with the original model developers (challenge 2).
-Similarly, when a modeler develops a new model, she can immediately  validate it against all the tests it is capable of taking.
-If a statistician develops a better goodness-of-fit metric, it too can be turned into a collection of tests that use existing capabilities, and so the table can be regenerated automatically, as discussed above.
+As long as the interface between the test and the model remains the same, the scores for prior models can be recomputed entirely automatically, requiring no coordination with model developers (challenge 4).
+When a modeler develops a new model, she can   validate it against all the tests it is capable of taking.
+If a statistician develops a better goodness-of-fit metric, it too can existing capabilities, and so a comprehensive score table can be generated easily, as discussed above.
 If a scientist is not interested in certain tests, she can locally remove them to recalibrate her view of the relative merits of different models (challenge 5).
 
-Once satisfied, these scientists might submit a \emph{pull request} to the central repository (either separately or in conjunction with the submission of a conventional paper).
+Once satisfied with their contributions to the repository, these scientists might submit a \emph{pull request} to the central repository (either separately or together with the submission of a conventional paper).
 Anonymous reviewers tasked by the ``editors'' of the repository would then subject the pull request to appropriate peer review before committing it to the ``academic record''.
-A model that performs well across tests in a collaboratively developed suite could more easily claim (e.g.
+A model that performs well across tests in a collaboratively developed suite could more directly and believably claim (e.g.
 to reviewers) to be a coherent, valid model of, for example, Saturn's motion.
@@
 
-To help help scientists discover these repositories and agree on which ones are central, we are developing a collaborative portal called \textit{SciDash} (\url{http://scidash.org/}).
-It is essentially a wiki pointing scientists to the $X$\verbx{unit}and $X$\verbx{suite} repositories designated as ``primary'' by early adopters in their research areas (who serve as \emph{de facto} initial editors as well).
-SciDash also supports loading IPython notebooks containing generated tables stored in these repositories into the browser directly so that score tables can be explored without going to the trouble of cloning the repository locally.
+To help scientists discover and coordinate control over these repositories, we are developing a collaborative portal called \textit{SciDash} (\url{http://scidash.org/}).
+It is essentially a wiki pointing scientists to the common and suite repositories designated as ``primary'' by early adopters in their research areas (who serve as \emph{de facto} initial editors as well).
+SciDash also supports loading IPython notebooks containing score tables stored in these repositories into the browser directly, avoiding the need to clone the repository locally.
 To modify the suite (e.g.
-by adding the model a scientist is developing to it), a scientist can simply fork the repository by clicking a button at the top of such a view.
+to add the model a scientist is developing to it), a scientist can simply fork the repository by clicking a button at the top of this view.
 We are exploring ways of loading forks directly into a cloud-based IPython notebook system, so that the complexities of Git and installation of Python and the packages required by the repository are left behind the scenes (challenge 4).
-SciDash discovers public forks of repositories that it is indexing and lists them to help scientists get a better sense of who is working on the problems they are interested in and what partial results are publicly available.
+SciDash discovers public forks of repositories that it is indexing and lists them to help scientists get a better sense of who is working on the problems they are interested in and what partial results are publicly available, fostering collaboration.
 
-TODO: How to get adoption?
-Run contests -- dash double meaning.
-Also ask people directly (worked for Scholarpedia...) We also plan to use SciUnit to implement competitions among classes of models or techniques in particular subfields.
-Typically such competitions use \emph{ad hoc} infrastructure.
-A common framework could be used to make designing and continuously operating such contests more feasible.
+As with many pieces of cyberinfrastructure, early adopters may benefit less than late adopters due to network effects.
+To incentivize adoption, we look to an increasingly common scientific practice: modeling competitions.
+Many research communities use these to periodically determine the state of the art, resulting in a form of review paper.
+Funding sources often include small monetary incentives.
+%For example, the quantitative single neuron modeling competition operated over the course of two years  \cite{qsnmc}. 
+   SciDash could be used to run these kinds of competitions on a continuous basis atop less \textit{ad hoc} infrastructure (and thus require less effort and central planning from organizers) and we are actively organizing such competitions.
 
@@
 
-\section{Discussion}\label{discussion}
-The design just described has been purposefully left as simple as possible, in pursuit of challenge 4.
-Despite its simple design, it captures the essential elements of the scientific model validation process and satisfies the four challenges we laid out.
-TODO: SciUnit is essentially an embedded DSL that captures precisely the standard workflow of model validation.
-SciUnit can be used by individual scientists to organize their workflows, but because science, and in particular, model validation is a collaborative process, we have also described the intended use of SciUnit within a collaborative workflow, mediated by a social coding tool like Github.
+\vspace{5px}
+\section{Discussion \& Future Directions}\label{discussion}
+Instead of attempting to use an existing general-purpose testing infrastructure, we have designed what is essentially a domain-specific language that captures the scientific model validation process in more familiar terms (challenge 6).
+SciUnit can be used by individual scientists to organize their workflows and also within a loosely collaborative workflow, mediated by a social coding tool like Github and an easy-to-use portal, SciDash, based on the IPython system \cite{perez2007ipython}.
 
@@
 
-While we discuss a toy example based on planetary movement here, we have applied this framework to more realistic problems in neurobiology in a project called NeuronUnit (http://github.com/scidash/neuronunit).
-We have used this as part of a collaboration with a large-scale neuroinformatics project (http://www.opensourcebrain.org), whose developers are enthusiastic about a practical pipeline for testing the dozens of models they currently maintain.
-TODO: More SE-research-friendly future work discussion Much future work remains to be done to investigate whether these tools are truly usable and useful to scientific communities, to further develop the community workflow and infrastructure and to develop a range of realistic case studies.
-Nevertheless, we believe that identifying the synergies between testing practices and model validation, and describing the basic tooling, represents a novel contribution to the literature on software testing.
-The core SciUnit framework has been fully developed and is available at http://sciunit.scidash.org/.
-SciDash is under active development, and is currently capable of basic forms of the operations described in Sec.
-3.
+While we discuss a simple example based on planetary movement here, we have applied this framework to more realistic problems in neurobiology in collaboration with a large-scale informatics project, OpenSourceBrain, whose developers are enthusiastic about developing a practical pipeline for precisely characterizing the scope and validity of the dozens of models they currently maintain (case study in preparation).
+The tools we describe could also be used for a variety of other scientific disciplines, including within software engineering to validate quantitative models of, for example, software cost or component reusability.
 
+This paper identifies potential  synergies between software development and model validation practices, and provides a proof of concept to argue that a test-driven development methodology could help improve scientific workflows.
+The modern scientific model validation process has not yet been extensively studied by  the research community.
+Precisely characterizing how scientists today attempt to answer the questions in the introduction, and evaluating the effectiveness of tools like the ones we propose, are likely fruitful avenues for future empirical research.
+
+The core SciUnit framework has been fully developed and is available at http://sciunit.scidash.org/. SciDash is under active development as of this writing.
+\vspace{5px}
  %In particular, we consider modeling competitions in science as a form of test-driven validation.
@@
 
-\section*{Acknowledgements}
-We thank Sharon Crook, Shreejoy Tripathy and Padraig Gleeson for their helpful discussions.  
-The work was supported in part by NIMH grant R01MH081905. 
+\section{ACKNOWLEDGEMENTS}
+We thank Sharon Crook, Shreejoy Tripathy and Padraig Gleeson for their many helpful discussions.  
+This work was supported in part by NIMH grant R01MH081905. 
 The content is solely the responsibility of the authors and does not necessarily represent the official views of the NIH.
-
+\vspace{5px}
 % conference papers do not normally have an appendix
@@
 % (used to reserve space for the reference number labels box)
+%\renewcommand\bibsection{\subsection{\refname}}
+\renewcommand\refname{New Title}
 \bibliographystyle{abbrv}

### 2014-03-14 sciunit-icse14/sciunit-icse14.tex
ICSE14: Final draft

@@
 
-While we discuss a simple example based on planetary movement here, we have applied this framework to more realistic problems in neurobiology in collaboration with a large-scale informatics project, OpenSourceBrain, whose developers are enthusiastic about developing a practical pipeline for precisely characterizing the scope and validity of the dozens of models they currently maintain (case study in preparation).
-The tools we describe could also be used for a variety of other scientific disciplines, including within software engineering to validate quantitative models of, for example, software cost or component reusability.
+While we discuss a simple example based on planetary movement here, we have applied this framework to more realistic problems in neurobiology in collaboration with a large-scale informatics project, OpenSourceBrain, whose developers are enthusiastic about precisely characterizing the scope and validity of the dozens of models they currently maintain (case study in preparation).
+The tools we describe could also be used for a variety of other scientific disciplines, including within software engineering to validate quantitative models of, for example, software cost or reusability.
 

### 2014-03-15 sciunit-icse14/sciunit-icse14.tex
ICSE14: Final draft rtpos
ICSE14: Fixed some final typos

@@
 Today, a scientist demonstrates the validity of a model by making an argument in a paper and submitting it for peer review, a process comparable to \emph{code review} in software engineering.
-While human review helps to ensure that contributions meet high-level goals, software engineers typically supplement it with \emph{unit testing} to get a more complete picture of the status of a project, particularly when it is complex and involves many contributors.
+While human review helps to ensure that contributions meet high-level goals, software engineers typically supplement it with \emph{unit testing} to get a more complete picture of the status of a project.%, particularly one involving many contributors.
 
@@
 Scientific communities differ from software communities in several key ways, however.
-In this paper, we introduce \emph{SciUnit}, a framework for test-driven scientific model validation and outline how SciUnit, supported by new and existing collaborative infrastructure, could be  integrated into the modern scientific workflow.
+In this paper, we introduce \emph{SciUnit}, a framework for test-driven scientific model validation, and outline how, supported by new and existing collaborative infrastructure, it could be  integrated into the modern scientific process.
 \end{abstract}
@@
 
-Quantitative models are today validated by \emph{peer review}.
+Today, quantitative models are validated by \emph{peer review}.
 For a model to be accepted by a scientific community, its advocates must submit a paper that describes how it works and provides evidence that it makes more accurate predictions than previous models (or that it makes a desirable tradeoff  between accuracy and complexity) \cite{box1987empirical}.
-Other members of the relevant community are then tasked with ensuring that validity was measured properly and that all relevant data and competing models were adequately considered, drawing on knowledge of statistical methods and the prior literature.
+Other members of the relevant community are then tasked with ensuring that validity was measured correctly and that all relevant data and competing models were adequately considered, drawing on knowledge of statistical methods and the prior literature.
 Publishing is a primary motivator for most scientists \cite{howison2011scientific}.
@@
 Each paper is narrowly focused, often considering just one model or dataset, and is frozen in time, so it does not consider the latest models, data or statistical methods.
-Discovering, precisely characterizing and comparing models to discover the state of the art and find open modeling problems can require an encyclopedic knowledge of the literature, as can finding all data relevant to a model.
-Senior scientists often attempt to fill this need by publishing review papers, but in many areas, the number of publications generated every year can be overwhelming \cite{jinha_article_2010}, and comprehensive reviews of a particular area are published infrequently.
-Statisticians often complain that scientists are not following best practices and that community standards evolve too slowly because a canonical paper or popular review used outdated methods.
-Furthermore, if the literature simply doesn't address an important question of validity, a researcher might need to reimplement an existing model in order to evaluate it.
+Discovering, precisely characterizing and comparing models to determine the state of the art and find open modeling problems can require an encyclopedic knowledge of the literature, as can finding all data relevant to a model.
+Senior scientists often attempt to fill these needs by publishing review papers, but in many areas, the number of publications generated every year can be overwhelming \cite{jinha_article_2010}, and comprehensive reviews of a particular area are published infrequently.
+Statisticians often complain that scientists are not following best practices and that community standards evolve too slowly because a canonical paper or popular review relied on outdated methods.
+Furthermore, if the literature simply doesn't address an important question of validity (e.g.
+if space was limited), a researcher might need to reimplement an existing model from scratch to attempt to address it.
 
@@
 \ref{scidash}.
-The overall goal of this work is to help scientists generate and examine tables like the one central to Figure 1, where the relative validity of a set of models having a common scope can be determined by examining scores produced by a suite of validation tests constructed from experimental data.
-We discuss adoption strategies and directions for future research into scientific model validation practices in Sec. \ref{discussion}.
+The overall goal of this work is to help scientists generate tables like the one central to Figure 1, where the relative validity of a set of models having a common scope can be determined by examining scores produced by a suite of validation tests constructed from experimental data.
+We discuss adoption strategies and conclude in Sec. \ref{discussion} with directions for future research into scientific model validation practices.
 %Anecdotally, one of the greatest fears of new scientists is that a problem that they are working on has already been solved and published.
@@
 This illustrates a key difference between unit testing, which would simply produce a boolean result, and our conception of scientific validation testing (challenge 1).
-A score class must induce an ordering, so that a table like that shown in Figure 1 can be sorted along its columns, and it can  specify a normalization scheme so the cells can be color-coded.
+A score class must induce an ordering, so that a table like that shown in Figure 1 can be sorted along its columns.
+A score class can also specify a normalization scheme so the cells of the table can be color-coded.
 %It also makes it easier to cache predictions generated by large-scale models so that they can be used by multiple tests.
 
-Te \verbx{judge} method of a test can be invoked to compute a score for a single model: 
+The \verbx{judge} method of a test can be invoked to compute a score for a single model: 
 \vspace{-4px}
@@
 \vspace{-5px}
-This method proceeds by first checking that the provided model implements all required capabilities before calling the \verbx{generate_prediction} method followed immediately by the \verbx{compute_score} method.
-A reference to the test,   model, observation, prediction and other related data the test provides (none here) are available as attributes of the score.
+This method first checks that the model implements all required capabilities, then calls the \verbx{generate_prediction} method, then passes the prediction to the \verbx{compute_score} method.
+A reference to the test,   model, observation, prediction and other related data the test provides (none here) are added as attributes of the score before it is returned.
 
@@
 When a modeler develops a new model, she can   validate it against all the tests it is capable of taking.
-If a statistician develops a better goodness-of-fit metric, it too can existing capabilities, and so a comprehensive score table can be generated easily, as discussed above.
+If a statistician develops a better goodness-of-fit metric, she too can use existing capabilities and indeed inherit prediction extraction logic directly, as discussed above.
 If a scientist is not interested in certain tests, she can locally remove them to recalibrate her view of the relative merits of different models (challenge 5).
 
-Once satisfied with their contributions to the repository, these scientists might submit a \emph{pull request} to the central repository (either separately or together with the submission of a conventional paper).
-Anonymous reviewers tasked by the ``editors'' of the repository would then subject the pull request to appropriate peer review before committing it to the ``academic record''.
-A model that performs well across tests in a collaboratively developed suite could more directly and believably claim (e.g.
+If the changes to the repository appear to represent a contribution, these scientists might submit a \emph{pull request} to the central repository (either separately or together with the submission of a conventional paper).
+Anonymous reviewers tasked by the ``editors'' of the repository would then subject the pull request to appropriate peer review before committing it to the ``permanent record''.
+A model that performs well across tests in a collaboratively developed suite could more objectively claim (e.g.
 to reviewers) to be a coherent, valid model of, for example, Saturn's motion.
@@
 
-To help scientists discover and coordinate control over these repositories, we are developing a collaborative portal called \textit{SciDash} (\url{http://scidash.org/}).
-It is essentially a wiki pointing scientists to the common and suite repositories designated as ``primary'' by early adopters in their research areas (who serve as \emph{de facto} initial editors as well).
-SciDash also supports loading IPython notebooks containing score tables stored in these repositories into the browser directly, avoiding the need to clone the repository locally.
+To help scientists discover and coordinate development of these repositories, we are developing a collaborative portal called \textit{SciDash} (\url{http://scidash.org/}).
+It is essentially a wiki pointing scientists to the repositories designated as ``primary'' by early adopters in their research areas (who may serve as \emph{de facto} initial editors as well).
+SciDash also supports loading IPython notebooks containing score tables stored in these repositories into the browser, avoiding the need to clone the repository locally.
 To modify the suite (e.g.
 to add the model a scientist is developing to it), a scientist can simply fork the repository by clicking a button at the top of this view.
-We are exploring ways of loading forks directly into a cloud-based IPython notebook system, so that the complexities of Git and installation of Python and the packages required by the repository are left behind the scenes (challenge 4).
-SciDash discovers public forks of repositories that it is indexing and lists them to help scientists get a better sense of who is working on the problems they are interested in and what partial results are publicly available, fostering collaboration.
+We are exploring ways of loading forks directly into a cloud-based IPython notebook system, so that the complexities of Git, Python setup and package management can be left behind the scenes (challenge 6).
+SciDash lists public forks of repositories that it is indexing to help scientists get a better sense of who is working on the problems they are interested in and what partial results are publicly available, fostering collaboration.
 
@@
 %For example, the quantitative single neuron modeling competition operated over the course of two years  \cite{qsnmc}. 
-   SciDash could be used to run these kinds of competitions on a continuous basis atop less \textit{ad hoc} infrastructure (and thus require less effort and central planning from organizers) and we are actively organizing such competitions.
+   SciDash could be used to run these kinds of competitions on a continuous basis atop less \textit{ad hoc} infrastructure (and thus require less effort and central planning from organizers) and we are actively recruiting organizers for such competitions.
 
@@
 Instead of attempting to use an existing general-purpose testing infrastructure, we have designed what is essentially a domain-specific language that captures the scientific model validation process in more familiar terms (challenge 6).
-SciUnit can be used by individual scientists to organize their workflows and also within a loosely collaborative workflow, mediated by a social coding tool like Github and an easy-to-use portal, SciDash, based on the IPython system \cite{perez2007ipython}.
+SciUnit can be used by individual scientists to organize their workflows using IPython notebooks \cite{perez2007ipython} and also within a loosely collaborative workflow, mediated by a social coding tool like Github and a simple portal, SciDash.
 
@@
 While we discuss a simple example based on planetary movement here, we have applied this framework to more realistic problems in neurobiology in collaboration with a large-scale informatics project, OpenSourceBrain, whose developers are enthusiastic about precisely characterizing the scope and validity of the dozens of models they currently maintain (case study in preparation).
-The tools we describe could also be used for a variety of other scientific disciplines, including within software engineering to validate quantitative models of, for example, software cost or reusability.
+The tools we describe could also be used for a variety of other scientific disciplines, including within software engineering to validate predictive models of, for example, development time or code reuse.
 
-This paper identifies potential  synergies between software development and model validation practices, and provides a proof of concept to argue that a test-driven development methodology could help improve scientific workflows.
-The modern scientific model validation process has not yet been extensively studied by  the research community.
-Precisely characterizing how scientists today attempt to answer the questions in the introduction, and evaluating the effectiveness of tools like the ones we propose, are likely fruitful avenues for future empirical research.
+This paper identifies potential  synergies between software development and model validation practices, and sketches a sociotechnical solution to several challenges to argue that a test-driven development methodology could help improve the scientific process.
+The scientific model validation process has not yet been extensively studied, so precisely characterizing how scientists attempt to answer the questions in the introduction today, and evaluating how effective tools like the ones we propose are in improving their ability to do so, are likely fruitful avenues for future empirical research.
 
@@
 \section{ACKNOWLEDGEMENTS}
-We thank Sharon Crook, Shreejoy Tripathy and Padraig Gleeson for their many helpful discussions.  
-This work was supported in part by NIMH grant R01MH081905. 
+We thank Sharon Crook, Shreejoy Tripathy and Padraig Gleeson for their many helpful discussions, and the anonymous referees and several friends for their useful comments.  
+RG was supported in part by NIMH grant R01MH081905. 
 The content is solely the responsibility of the authors and does not necessarily represent the official views of the NIH.

### 2014-03-15 sciunit-icse14/sciunit-icse14.tex
ICSE14 final

@@
 % can use linebreaks \\ within to get better formatting as desired
-\title{Collaborative Infrastructure for Test-Driven\\Scientific Model Validation}
+\title{Collaborative Infrastructure for Test-Driven \\Scientific Model Validation}
 
@@
 \caption{\small Tests are derived from data and models are derived from scientific theories, either directly or through existing data and model repositories.
-The score table summarizes the performance of a collection of models against a suite of tests.
+The score table (center) summarizes the performance of a collection of models against a suite of tests.
 Tests and models are initialized, executed and visualized using \texttt{sciunit} in an IPython notebook stored inside a \emph{suite repository} (e.g.
@@
 Github).
-SciDash  discovers and organizes  suite repositories and provides interactive views of the notebooks they contain.
-Common testing code, capabilities and bridges to external model and data repositories are also collaboratively developed (e.g.
+SciDash  discovers and organizes  suite repositories and provides utilities for viewing them online without needing to create a local copy.
+Logic common across research areas is also collaboratively developed (e.g.
 \texttt{cosmounit}).}
@@
 To incentivize adoption, we look to an increasingly common scientific practice: modeling competitions.
-Many research communities use these to periodically determine the state of the art, resulting in a form of review paper.
+Many research communities use these to periodically determine the state of the art, resulting in a form of review paper (e.g.
+\cite{jolivet_quantitative_2008}).
 Funding sources often include small monetary incentives.
@@
 \section{Discussion \& Future Directions}\label{discussion}
-Instead of attempting to use an existing general-purpose testing infrastructure, we have designed what is essentially a domain-specific language that captures the scientific model validation process in more familiar terms (challenge 6).
-SciUnit can be used by individual scientists to organize their workflows using IPython notebooks \cite{perez2007ipython} and also within a loosely collaborative workflow, mediated by a social coding tool like Github and a simple portal, SciDash.
+Instead of attempting to use an existing general-purpose testing framework (e.g.
+JUnit or PyUnit), we have designed what is essentially a domain-specific language that captures the scientific model validation process in more familiar terms (in view of challenge 6).% SciUnit can be used by individual scientists to organize their workflows using IPython notebooks \cite{perez2007ipython} and also within a loosely collaborative workflow, mediated by a social coding tool like Github and a simple portal, SciDash.
 
@@
 
-While we discuss a simple example based on planetary movement here, we have applied this framework to more realistic problems in neurobiology in collaboration with a large-scale informatics project, OpenSourceBrain, whose developers are enthusiastic about precisely characterizing the scope and validity of the dozens of models they currently maintain (case study in preparation).
-The tools we describe could also be used for a variety of other scientific disciplines, including within software engineering to validate predictive models of, for example, development time or code reuse.
+While we discuss a simple example based on planetary movement here, we have applied this framework to more realistic problems in neurobiology in collaboration with the developers of two large-scale informatics projects, OpenSourceBrain and NeuroElectro, which serve as a source of models and data, respectively.
+Our tools serve, in essence, as a bridge between these resources.
+The tools we describe could also be used for a variety of other scientific disciplines, including within software engineering to validate predictive models of, for example, project duration or code reuse.
 
-This paper identifies potential  synergies between software development and model validation practices, and sketches a sociotechnical solution to several challenges to argue that a test-driven development methodology could help improve the scientific process.
-The scientific model validation process has not yet been extensively studied, so precisely characterizing how scientists attempt to answer the questions in the introduction today, and evaluating how effective tools like the ones we propose are in improving their ability to do so, are likely fruitful avenues for future empirical research.
+This work identifies new research directions in software engineering by developing connections between test-driven development \cite{beck2003} and the scientific model validation process, which has not been as extensively studied.
+Characterizing how scientists today attempt to answer the questions given in the introduction, evaluating how well sociotechnical solutions like the ones we propose here could improve their ability to do so, and studying adoption barriers are likely to be fruitful avenues for future empirical research, and lead to a better understanding of both programmers and scientists.
 
@@
 \section{ACKNOWLEDGEMENTS}
-We thank Sharon Crook, Shreejoy Tripathy and Padraig Gleeson for their many helpful discussions, and the anonymous referees and several friends for their useful comments.  
+We thank Sharon Crook, Shreejoy Tripathy and Padraig Gleeson for their many helpful discussions, and anonymous referees and several friends for their careful readings.  
 RG was supported in part by NIMH grant R01MH081905. 

### 2014-03-15 sciunit-icse14/sciunit-icse14.tex
ICSE14 final final

@@
 Scientific communities differ from software communities in several key ways, however.
-In this paper, we introduce \emph{SciUnit}, a framework for test-driven scientific model validation, and outline how, supported by new and existing collaborative infrastructure, it could be  integrated into the modern scientific process.
+In this paper, we introduce \emph{SciUnit}, a framework for test-driven scientific model validation, and outline how, supported by new and existing collaborative infrastructure, it could integrate into the modern scientific process.
 \end{abstract}

### 2014-03-15 att-icfp14/att-icfp14.tex
removed stuff

@@
 parametric polymorphism), we must take a moment to reflect.
-In $\mathcal{L}\{\rightarrow\,\forall\}$, studied variously by Girard as System $\mathbf{F}$ \cite{girardF} and Reynolds as the polymorphic lambda calculus \cite{polylam}, it is known that sums, products, and inductive and co-inductive types can all be weakly encoded \cite{reynolds}\todo{what to cite for this?}.
+In $\mathcal{L}\{\rightarrow\,\forall\}$, studied variously by Girard as System $\mathbf{F}$ \cite{girard1971extension} and Reynolds as the polymorphic lambda calculus \cite{Reynolds94anintroduction}, it is known that sums, products, and inductive and co-inductive types can all be weakly defined.
 This means that we can \emph{translate} well-typed terms of a language like $\mathcal{L}\{\rightarrow\forall~\mathtt{nat} + \times\}$ to well-typed terms in $\mathcal{L}\{\rightarrow\forall\}$ in a manner that preserves their dynamic semantics.
-But Reynolds, in a remark that recalls the ``Turing tarpit'' of Perlis \cite{Perl82a}, reminds us that programming with the corresponding \emph{embedding}, particularly one where only the dynamic semantics are preserved, may be unwise \cite{Reynolds94anintroduction}:
+But Reynolds, in a remark that recalls the ``Turing tarpit'' of Perlis \cite{Perl82a}, reminds us that programming with the corresponding \emph{embedding} may be unwise \cite{Reynolds94anintroduction}:
 \begin{quote}
@@
 %Having established why a minimal language like $\mathcal{L}\{\rightarrow\forall\}$, while suitable as an internal language, needs to be extended with additional fragments before it is suitable for use as a human-facing external language,
-\todo{revise per title}A strong embedding, i.e.
-an isomorphism that preserves static reasoning principles, in a sense that we will make more precise as we go on, is generally far more difficult to establish than a weak embedding.
-For example, the aforementioned embeddings into $\mathcal{L}\{\rightarrow\forall\}$ do not preserve type disequality and other equational principles.
-Harder still is establishing a strong embedding that also supports an acceptably natural programming style (measured, to first approximation, by the amount of boilerplate code needed) and a reasonable cost semantics.
+A strong embedding, i.e.
+an isomorphism that preserves both static reasoning principles and dynamic behaviors, in a sense that we will make more precise as we go on, is generally far more difficult to establish than a weak embedding.
+For example, the aforementioned embeddings into $\mathcal{L}\{\rightarrow\forall\}$ do not preserve type disequality and some other equational reasoning principles (we will see less subtle issues soon).
+Establishing a strong embedding that is natural (as measured, to a first approximation, by the ratio of boilerplate code that must be manually generated in the embedding vs.
+in the fragment, and how likely one is to make a mistake when writing it) and that has a reasonable cost semantics is harder still.
 
@@
 %\compresslist
-\item General-purpose constructs continue to admit variants.
-There are many  variants of product types: records, records with functional record update, labeled tuples (records that specify a field ordering) and records with delegation are either awkward to strongly embed or cannot be.
+\item General-purpose constructs continue to evolve to better handle various usage scenarios.
+There are many  variants of product types: records, records with functional record update, labeled tuples (records that specify a field ordering) and record-like data structures with field delegation (of various sorts).
+These are  either awkward to strongly embed or cannot be.
 Similarly, sum types also admit many variants (e.g.
 various forms of open sums and object systems, seen in a certain light).
-Even something as seemingly simple as a type-safe \verb|sprintf| operator requires special support from a language, e.g.
-Ocaml \cite{ocaml-printf}.
+Even something as seemingly simple as a type-safe \verb|sprintf| operator requires special support from languages like Ocaml.
 \item Perhaps more interestingly, specialized type systems that enforce stronger invariants than general-purpose constructs are capable of enforcing are often developed by researchers.
-One need take only a brief  excursion through the literature to discover language extensions that support parallel programming \cite{a}\todo{citations}, concurrent programming \cite{cml}, distributed programming \cite{tom7}, dataflow programming \cite{reactiveml}, authenticated data structures \cite{popl13}, information flow security \cite{walker00, smith2001}, database queries \cite{db}, aliased references \cite{naden12}, network protocols \cite{sekar99}, units of measure \cite{keneddy} and many others.% All of these are implemented as dialects of existing languages, presumably because a strong encoding was not feasible.
+One need take only a brief  excursion through the literature to discover language extensions that support data parallel programming \cite{chakravarty2007data}, concurrency \cite{reppy1993concurrent}, distributed programming \cite{Murphy:2007:TDP:1793574.1793585}, dataflow programming \cite{mandel2005reactiveml}, authenticated data structures \cite{Miller:2014:ADS:2535838.2535851}, database queries \cite{Ohori:2011:MSM:2034773.2034815},  units of measure \cite{conf/cefp/Kennedy09} and many other domains.% All of these are implemented as dialects of existing languages, presumably because a strong encoding was not feasible.
 \item Safe and natural foreign function interfaces (FFIs) require enforcing the type system of the foreign language within the calling language.
@@
 %Safe FFIs generally require direct extensions to the language. 
-For example, MLj extended Standard ML with constructs for safely and naturally interfacing with Java \cite{mlj}.
+For example, MLj extended Standard ML with constructs for safely and naturally interfacing with Java \cite{benton1999interlanguage}.
 For any other language, including  others on the JVM like Scala and the many languages that might be accessible via a native FFI, there is no way to guarantee that language-specific invariants are statically maintained, and the interface is certainly far from natural.
@@
 Forming such a dialect is alarmingly non-trivial, even in the rare situation where a common framework has been used or the dialects are implemented within the same compiler, as these mechanisms do not guarantee that different combinations of individually sound dialects remain sound when combined.
-Metatheoretic and compiler correctness results can only be derived for the dialect \emph{resulting} from a language   composition operation, so in a diverse software ecosystem, fragment providers have little choice but to leave this task to clients (providing, perhaps, some informal guidelines or partial automation).
-Avoiding language composition is difficult, as interoperability between components of an application written in different dialects faces precisely the problems of item 3.% We will consider this further in our discussion of related work in Sec.
+Metatheoretic and compiler correctness results can only be derived for the dialect \emph{resulting} from a language   composition operation, so in a sufficiently diverse software ecosystem, dialect providers have little choice but to leave this task to clients (providing, perhaps, some informal guidelines or partial automation).
+Avoiding dialect composition is difficult, because interactions between components of an application written in different dialects can lead to precisely the problems of item 3.% We will consider this further in our discussion of related work in Sec.
 \ref{related-work}.
@@
 These are not the sorts of problems usually faced by library providers.
-Well-designed languages preclude the  possibility of conflict between two libraries and ensure that the semantics of one library  cannot be weakened by another by strictly enforcing abstraction barriers.
+Well-designed languages preclude the  possibility of ``link-time'' conflicts between libraries and ensure that the semantics of one library  cannot be weakened by another by strictly enforcing abstraction barriers.
 For example, a module declaring an abstract type in ML can rely on its representation invariants no matter which other modules are in use, so clients can assume that they will operate robustly in combination without needing to attend to burdensome ``link-time'' proof obligations.
@@
 \paragraph{Contributions}
-We draw inspiration from this approach to design a mechanism that makes it possible to declare new type and operator constructors and define, \emph{using type-level functions}, their static and dynamic semantics, the latter by translation to a fixed typed internal language.
-If a language fragment can be weakly defined in terms of the typed internal language and its statics are of a form that can be implemented by this protocol, it can be made strongly embeddable in the external language by introducing, in this way, the necessary constructors and type assignment logic.
+Drawing from this approach, we aim to introduce a mechanism that makes it possible to declare new type and operator constructors and define, \emph{using type-level functions}, their static and dynamic semantics, the latter by translation to a fixed typed internal language.
+If a language fragment can be weakly defined in terms of the typed internal language and its statics are of a general form consistent with this protocol, a  strong embedding in the external language can be constructed by simply introducing the necessary constructors and defining directly the necessary type assignment and translation logic.
 We call such an embedding an \emph{active embedding}.
 
-The semantics imposes checks on extensions to guarantee type safety and decidability of type assignment and equality.
-Moreover, an important class of lemmas that play a key role in  proofs that an extension enables strong embeddings of a language fragment can be derived in a suitable ``closed world'' (that is, in the traditional manner where one reasons inductively over a language composed of a minimal collection of fragments).
-The semantics guarantees that they will be \emph{conserved} in the  ``open world'' by enforcing a form of  representation independence at extension boundaries.
-This avoids the most fundamental problems of composing dialects and justifies the inclusion of the mechanism inside the language.
+The semantics imposes various checks to ensure type safety and decidability of type assignment and equality.
+Moreover, an important class of lemmas that play a key role in  proofs that a library enables strong embeddings of a language fragment can be derived in a suitable ``closed world'' (that is, in the traditional manner where one reasons inductively over a \emph{finite} collection of constructors).
+The semantics guarantees that they will be \emph{conserved} in the  ``open world'' by enforcing an abstraction barrier at extension boundaries.
+This avoids the most fundamental semantic problems that can arise when composing dialects and justifies the inclusion of the mechanism inside the language.
 
@@
 The external language begins with only the $\rightarrow$ fragment and the typed internal language is a simple variant of PCF that permits only weak embeddings of even simple fragments like the $\mathtt{nat}$ fragment of G\"odel's $\mathbf{T}$ or ML-style $n$-ary products.
-We implement these as extensions to explain the mechanism and motivate the constraints that the semantics imposes.
+We actively embed these into the external language to explain the mechanism and motivate the constraints that the semantics imposes.
 We then examine the semantics in greater detail and state and sketch the key points in the proofs of the metatheory in Sec.
 \ref{theory}.
-These two sections are the fundamental contributions of this work.
-
-While the core calculus addresses the issue of making a strong embedding possible, the embedding into the core calculus is quite clearly impractical syntactically.
+These two sections comprise the fundamental contributions of this work.
+
+While the core calculus addresses the issue of making a strong embedding possible, active embeddings into the core calculus are often impractical syntactically.
 To begin to address the issue of syntax, we develop in Sec.
@@
 While of surprising expressiveness given its minimality, @$\lambda$ itself is certainly not capable of admitting embeddings of the full variety of type system fragments enumerated earlier.
-Indeed, to guarantee safety and conservativity, our calculus assumes perhaps too much about the type system being encoded.
+Indeed, to guarantee safety and conservativity and because our intention in this paper is to cleanly expose the essence of this mechanism, our calculus makes perhaps too many assumptions about the type system being embedded.
 For example, it only admits fragments where the typing judgement looks essentially like the typing judgement of the $\rightarrow$ fragment, so new contexts cannot be introduced.
@@
 \ref{limitations}.
-These limitations do not appear fundamental to the approach, so we also outline in Sec.
-\ref{limitations} directions for future work.
-We conclude with comparisons to related work in Sec.
+These limitations do not appear fundamental, so we also outline directions for future work.
+We conclude by considering related work in Sec.
 \ref{related-work}.
@@
 \ref{grammar}.
-An example of a program defining type and operator constructors that can be used to strongly encode G\"odel's \textbf{T} in @$\lambda$ is given in Fig.
+An example of a program defining type and operator constructors that admit an active embedding of G\"odel's \textbf{T} into @$\lambda$ is given in Fig.
 \ref{nat}.
-We will discuss its semantics and how precisely the corresponding embedding, seen being used starting on line 16, works as we go on.
-Natural numbers can, of course, be strongly embedded in existing languages with a similar usage and asymptotic performance profile (up to function call overhead with an abstract type, for example).
-We will provide more sophisticated examples where this is less feasible later on (and discuss type abstraction itself in Sec.
+We will discuss its semantics and how precisely that embedding, seen being used starting on line 16, works as we go on.
+Natural numbers can, of course, be strongly embedded in existing languages, with a similar usage and asymptotic performance profile (up to function call overhead as an abstract type, for example).
+We will provide more sophisticated examples where this is less feasible later on (and note that type abstraction itself in Sec.
 \ref{limitations}).\todo{if i don't have time/space, take this out} %It consists of:
@@
 % 												\trepof{\tau} \pipe \dup{\tau}\\
-\text{abstract IL}	& \bar{\iota} & ::= & x \pipe \ifix{x}{\bar \sigma}{\bar \iota} 
+\text{translational IL}	& \bar{\iota} & ::= & x \pipe \ifix{x}{\bar \sigma}{\bar \iota} 
 	%\pipe \ilam{x}{\bar \sigma}{\bar \iota} \pipe \iapp{\bar \iota_1}{\bar \iota_2} 
@@
  & \bar{\sigma} & ::= & \darrow{\bar \sigma_1}{\bar \sigma_2} \pipe \cdots \pipe \dup{\tau} \pipe \trepof{\tau} \\
+%\text{abstracted} & \sabs & ::= & \darrow{\sabs_1}{\sabs_2} \pipe \cdots \pipe \sabsrep{\tau}
 											\\
@@
 \caption{\small An embedding of G\"odel's $\mathbf{T}$ in \atlam, used to define $plus$ and $two$ and calculate $plus~two~two$.
-The simple helper functions $\tvar{arity}n$ and $\tvar{ifeq}$ have return kind $\ksum{\kDen}{\kunit}$ and should be replaced by the definitions given in Fig.
-\ref{helpers}.
-We use \textsf{let} to bind both external and type-level variables for clarity of presentation.
+Type constructors (line 1) and types (line 16) are discussed in Sec.
+\ref{types}.
+Schemas (line 1) and operator constructors (lines 2-14, used on lines 17-20) are discussed starting in Sec.
+\ref{opcons}.
+We use \textsf{let} to bind both external and type-level variables as needed for clarity of presentation.
 We will discuss further syntactic issues in Sec.
@@
 The syntax for external terms contains three forms: variables, $\lambda$ terms, and a form for invoking user-defined operators, which we will discuss below.
-Compiling a program consists of first \emph{kind checking} its static declarations and type-level terms, $\tau$, which we will describe in Sec \ref{types} (Fig.
-\ref{kindprog}), then typechecking the external term and simultaneously {translating} it to a term, $\iota$, in the {typed internal language}, described thereafter (Fig.
-\ref{att}).
-We can thus specify the \emph{central compilation judgement} $\pkcompiles{\rho}{\iota}$ with one rule:
+Compiling a program consists of first \emph{kind checking} it then typechecking the external term and {translating} it to a term, $\iota$, in the {typed internal language}.
+These correspond to the premises of the \emph{central compilation judgement} $\pkcompiles{\rho}{\iota}$:
 \[
 \inferrule[p-compiles]{
-	\progOK{\emptyset}{\fvalCtx_0}{\rho}\\
+	\emptyset \vdash_{\fvalCtx_0} \rho\\
+%	\progOK{\emptyset}{\fvalCtx_0}{\rho}\\
 	\pcompiles{\fvalCtx_0}{\rho}{\iota}
@@
 \]
-The key judgement form in the calculus is the \emph{active typing judgement}  (Fig.
-\ref{att}, which we will describe in Secs.
-\ref{opcons}-\ref{repind}), relating an external term to a {type}, $\tau$, called its \emph{type assignment}, and an internal term, called its \emph{translation}, under \emph{typing context} $\Gamma$ and \emph{constructor context} $\Phi$:
+This anchors our exposition; we will describe how it is derived (i.e. how to write a compiler for @$\lambda$) in the following sections.
+
+The key judgement in the calculus is the \emph{active typing judgement}  (Fig.
+\ref{att}, which we describe starting in Sec.
+\ref{opcons}).
+It relates an external term, $e$, to a {type}, $\tau$, called its \emph{type assignment}, and an internal term, $\iota$, called its \emph{translation}, under \emph{typing context} $\Gamma$ and \emph{constructor context} $\Phi$:
 \[\ecompilesX{e}{\tau}{\iota}\]
 
-The typing context $\Gamma$ is essentially conventional, mapping variables to types and obeying standard structural rules (weakening, exchange and contraction; \cite{pfpl} describes the necessary background for this paper).
-The constructor context tracks user-defined type and operator constructors, the topics of the remainder of this section, and establishing its structural properties requires establishing the conservativity properties that we will discuss in Sec.
-\ref{theory}.% Weakening and exchange of the constructor context closely related to the issue of conservativity that we will return to.
+The typing context $\Gamma$ maps variables to types in essentially the conventional way (\cite{pfpl} contains the necessary background for this paper).
+The constructor context tracks user-defined type and operator constructors.
+% Weakening and exchange of the constructor context closely related to the issue of conservativity that we will return to.
 Each constructor is identified by name, so there is no analog to contraction.
-
-There is no separate operational semantics for the external language; the dynamic behavior of an external term is determined by its translation to the internal language (which might have a more conventional operational semantics).
-This form of semantics can be considered as a lifting into the language specification of the first stage of a type-directed compiler like the TIL compiler for Standard ML \cite{TIL} and has some parallels to the Harper-Stone semantics for Standard ML, where external terms were also given meaning by elaboration to an IL \cite{harper-stone}\todo{citations}.
+There is no separate operational semantics for the external language.
+Instead, the dynamic behavior of an external term is determined by its translation to the internal language, which has a more conventional operational semantics.
+This form of semantics can be seen as lifting into the language specification the first stage of a type-directed compiler like the TIL compiler for Standard ML \cite{tarditi+:til-OLD} and has some parallels to the Harper-Stone semantics for Standard ML, where external terms were also given meaning by elaboration from the EL to an IL \cite{Harper00atype-theoretic}.
 %
@@
 In practice, the internal language could be any typed  language with a specification for which type safety and decidability of typechecking have been satisfyingly determined.
-Constraints that must be maintained for all well-typed terms, independent of the constructor context (e.g.
-guarantees that out-of-bounds access to memory never occurs), must be maintained by the internal type system and program performance is ultimately limited by the internal language and downstream compilation stages that we do not here consider (safely extending downstream compilation has been discussed in previous work, e.g.
-\cite{compexts}).
-
-\subsection{Type Constructors and Types (Fig. \ref{nat}, lines 1 and 16)}\label{types}
+The internal type system serves as a ``floor'': guarantees that must hold for \emph{all} well-typed terms, independent of the constructor context (e.g.
+that out-of-bounds access to memory never occurs), must be maintained by the internal type system.
+User-defined constructors can (only) enforce invariants stronger than those the internal type system maintains.
+Performance is also ultimately limited by the internal language and downstream compilation stages that we do not here consider (safe compiler extension has been discussed in previous work, e.g.
+\cite{conf/pldi/TatlockL10}).
+
+\subsection{Type Constructors and Types}\label{types}
 \begin{figure}[t]
@@
 $\fbox{\inferrule{}{\progOKX{\progsort}}}$
+~~~$\tvarCtx ::= \emptyctx \pipe \tvarCtxX{t}{\kappa}$
 ~~~$\fvalCtx ::= \emptyctx \pipe \fvalCtxX{\fvalDf}$
-~~~$\tvarCtx ::= \emptyctx \pipe \tvarCtxX{t}{\kappa}$
 %$\fCtx ::= \Sigma_0	 \pipe \fvalCtxX$
@@
 \end{mathpar}
-\caption{\small Kinding for programs, operator constructors and external terms.
-The kinding context $\tvarCtx$ maps type variables to kinds (we do not specify let binding of type-level variables at any level in the core, but this would be a simple addition).
+\caption{\small Kinding for type and operator constructors and external terms, and equality kinds (see text).
 Type-level terms stored in the constructor context are not needed during kinding, indicated using a dash.}
@@
 $\fbox{\inferrule{}{\tKindX{\tau}{\kappa}}}$
+%~~~$\chi ::= \cdot \pipe \checkmark$
 \begin{mathpar}
@@
 \inferrule[k-d-i-checked]{
-	\tKindX{\tau}{\kTypeBlur}\\\\
+	\tKind{\Delta}{\fvalCtx}{\tau}{\kTypeBlur}\\\\
 	\isItermX{\ibar}
 }{
-	\tKindX{\tden{\ibar}{\tau}^\checkmark}{\kDen}
-}
-
+	\tKind{\Delta}{\fvalCtx}{\tden{\ibar}{\tau}^\checkmark}{\kDen}
+}
+~~~~~~~
 \inferrule[k-d-e]{
@@
 }{
-	\tKindX{\ttypeof{\tau}}{\kDen}
+	\tKindX{\ttypeof{\tau}}{\kTypeBlur}
 }
@@
 }
-
+~~~~~~
 \inferrule[k-i-fix]{
-	\isItypeX{\sbar}\\\\
+	\isItypeX{\sbar}\\
 	\isItermX{\ibar}
@@
 }
-
+~~~~~~
 \inferrule[k-i-transof]{
@@
 
-\text{\color{gray} (omitted forms are treated recursively, as exemplified by \textsc{k-i-fix})}
+\text{\color{gray} (omitted rules are trivially recursive, like \textsc{k-i-fix})}
 %\inferrule[iterm-dereify-kinding]{
@@
 %
-\inferrule[k-s-parr]{
-	\isItypeX{\sigma_1}\\\\
-	\isItypeX{\sigma_2}
-}{
-	\isItypeX{\darrow{\sigma_1}{\sigma_2}}
-}
-~~~~~~
 \inferrule[k-s-unquote]{
@@
 }
-~~~~~~
+
 \inferrule[k-s-repof]{
@@
 
-\text{\color{gray} (omitted forms are treated recursively, as exemplified by \textsc{k-i-parr})}
+\text{\color{gray} (omitted forms are trivially recursive)}
 \end{mathpar}
-\caption{\small Kinding for type-level terms. Checked derivates should only be constructed by the semantics (a restriction that can be enforced by a parser).}
+\caption{\small Kinding for type-level terms.
+The kinding context $\tvarCtx$ maps type variables to kinds.
+Checked derivates (introduced by rule \textsc{k-d-i-checked}, discussed in Sec.
+\ref{repcon}) should only be constructed by the semantics (this can be enforced by a parser, so we will assume it in the metatheory).}
 \label{tlkind}
-%\vspace{-15px}
+\vspace{-10px}
 \end{figure} 
-In most languages, types are formed by applying one of a fixed collection of \emph{type constructors} to zero or more \emph{indices}.
+In most languages, types are formed by applying one of a collection of \emph{type constructors} to zero or more \emph{indices}.
 In @$\lambda$, the situation is notionally similar.
@@
 Specifically, type-level terms, $\tau$, themselves form a typed lambda calculus.
-The classifiers of type-level terms are, following convention, called \emph{kinds}, $\kappa$, to distinguish them from  \emph{types}.
-Types are  type-level values of kind $\kTypeBlur$, introduced as just described.
-The kind $\kTypeBlur$ also has an elimination form, $\tfamcase{\tau}{Tycon}{x}{\tau_1}{\tau_2}$, allowing the extraction of a type index by case analysis against an available type constructor.
-To a first approximation, we are treating type constructors as constructors of a built-in open datatype at the type-level.
-Like open datatypes, there is no notion of exhaustiveness and the default case is required for totality.
-%We will write the kind of types as $\kTypeBlur$, though it is also written $\star$ or \verb|Type| in various similarly structured languages (see Sec. \ref{related-work}). 
+The classifiers of type-level terms are called \emph{kinds}, $\kappa$, to distinguish them from  \emph{types}.
 The kinding rules are specified in Fig.
 \ref{kindprog} and Fig.
-\ref{tlkind} and the reader is encouraged to refer to these as we continue.% Rather than there being a fixed set of type constructors, we allow the programmer to declare new type  constructors, and give the static and dynamic semantics of their associated operators, by writing type-level functions.
+\ref{tlkind}.
+Types are  type-level values of kind $\kTypeBlur$, introduced as just described.
+The kind $\kTypeBlur$ also has an elimination form, $\tfamcase{\tau}{Tycon}{x}{\tau_1}{\tau_2}$, allowing the extraction of a type index by case analysis against a contextually-available type constructor.
+To a first approximation, one might think of type constructors as constructors of a built-in open datatype, $\kTypeBlur$, at the type-level.
+Like open datatypes, there is no notion of exhaustiveness so the default case is required for totality.
+%We will write the kind of types as $\kTypeBlur$, though it is also written $\star$ or \verb|Type| in various similarly structured languages (see Sec. \ref{related-work}). 
+% Rather than there being a fixed set of type constructors, we allow the programmer to declare new type  constructors, and give the static and dynamic semantics of their associated operators, by writing type-level functions.
 In the semantics for this calculus, our kind system combined with techniques borrowed from the typed compilation literature and a form of type abstraction allow us to prove strong type safety, decidability and conservativity theorems.
@@
 by unit kind, $\kunit$, so there is only one natural number type, $\ttype{Nat}{\tunit}$, but we will show examples of type constructors that are indexed in more interesting ways in later portions of this work.
-For example, $\fvar{Tuple}$ has index kind $\klist{\kTypeBlur}$ and $\fvar{LabeledTuple}$ has index kind $\klist{\kpair{\klabel}{\kTypeBlur}}$.
+For example, $\fvar{LabeledTuple}$ has index kind $\klist{\kpair{\klabel}{\kTypeBlur}}$.
 The type constructor $\fvar{Arrow}$ is included in the initial constructor context, $\fvalCtx_0$, shown in Fig.
 \ref{arrow}, and has index kind $\kpair{\kTypeBlur}{\kTypeBlur}$.
-We will discuss it further below.
+
+ Type constructors are not first-class; they do not themselves have arrow kind as in some kind systems (e.g.
+\cite{watkins2008specifying}; Ch.
+22 of \emph{PFPL} describes a related system \cite{pfpl}).
+The type-level language does, however, include total functions of conventional arrow kind, written $\karrow{\kappa_1}{\kappa_2}$.
+Type constructor application can be wrapped in a type-level function to emulate a first-class or uncurried version of a type constructor for convenience (indeed, such a wrapper could be generated automatically, though we do not do so).
 
@@
 Going further, we ensure that deciding type equivalence requires only checking for syntactic equality after normalization by imposing the restriction that equivalence at a type constructor's index kind must be decidable in this way.
-All combinations of kinds introduced thus far  have this property and this is sufficient for many interesting examples.
 Our treatment of equivalence in the type-level language is thus quite similar to the treatment of term-level equality using ``equality types'' in a language like Standard ML.
@@
 in \tvar{ifeq}, Fig.
-\ref{helpers}, discussed below).% as maintaining the metatheoretic guarantee that typing respects type equivalence would impose a substantial burden in such a setting.% (a na\"ive approach to this would impose non-trivial extrinsic proof obligations onto extension developers that, unlike in others in this thesis, could threaten type safety).
-
-Type constructors are not first-class; they do not themselves have arrow kind as in some kind systems (e.g.
-\cite{Watkins08}\todo{cite}; Ch.
-22 of \emph{PFPL} describes a related system \cite{pfpl}).
-The type-level language does, however, include total functions of conventional arrow kind, written $\karrow{\kappa_1}{\kappa_2}$.
-Type constructor application can be wrapped in a type-level function to emulate a first-class type constructor (indeed, such a wrapper could be generated automatically, though we avoid this for simplicity in our semantics).
-Equivalence at arrow kind does not meet our criteria, so type-level functions cannot appear within type indices.
+\ref{helpers}, discussed below).
+Equivalence at arrow kind is not decidable by our criteria, so type-level functions cannot appear within type indices.
 This also prevents general recursion from arising at the type level.
-Without this restriction, a type-level function taking a type as an argument could ``smuggle in'' a self reference as a type index and extract it via case analysis (continuing our analogy to open datatypes, this is closely related to the positivity condition for inductive datatypes in total languages).
-
-Every type constructor also defines a \emph{schema}, a type-level function that associates with every index an internal type.
-The \emph{representation type} of a type like $\ttype{Nat}{\tunit}$ is then determined by applying its constructor's schema to its index, here resulting in $\dint$.
+Without this restriction, a type-level function taking a type as an argument could ``smuggle in'' a self reference as a type index, extracting it via case analysis (continuing our analogy to open datatypes, this is closely related to the positivity condition for inductive datatypes in total functional languages like Coq).% as maintaining the metatheoretic guarantee that typing respects type equivalence would impose a substantial burden in such a setting.% (a na\"ive approach to this would impose non-trivial extrinsic proof obligations onto extension developers that, unlike in others in this thesis, could threaten type safety).
+
+Every type constructor also defines a \emph{schema}, a type-level function that associates with every type an internal \emph{representation type}.
 We will return to this in Sec.
@@
 \begin{array}{rcl}
-\tvar{tyerr} & := & \tinr{\kDen}{\tunit}\\
-\tvar{tyasgn} & := & (\tlam{d}{\kDen}{\tinl{\kunit}{\tvar{d}}})\\
+\tvar{failure} & := & \tinr{\kDen}{\tunit}\\
+\tvar{ok} & := & (\tlam{d}{\kDen}{\tinl{\kunit}{\tvar{d}}})\\
 \tvar{ifeq} & := & (\tlam{t1}{\kTypeBlur}{\tlam{t2}{\kTypeBlur}{\tlam{d}{\kDen}{
-	\\& & \quad \tifeq{\tvar{t1}}{\tvar{t2}}{\kTypeBlur}{\tvar{tyasgn}~\tvar{d}}{\tvar{tyerr}}\\
-\tvar{arity0} & := & (\tlam{a}{\klist{\kDen}}{\tlam{d}{(\ksum{\kDen}{\kunit})}{\tfold{\tvar{a}}{\tvar{d}}{\_}{\_}{\_}{\tvar{tyerr}}}})\\
+	\\& & \quad \tifeq{\tvar{t1}}{\tvar{t2}}{\kTypeBlur}{\tvar{ok}~\tvar{d}}{\tvar{failure}}\\
+\tvar{arity0} & := & (\tlam{a}{\klist{\kDen}}{\tlam{d}{(\ksum{\kDen}{\kunit})}{\tfold{\tvar{a}}{\tvar{d}}{\_}{\_}{\_}{\tvar{failure}}}})\\
 \tvar{arity1} & := & (\tlam{a}{\klist{\kDen}}{\tlam{k}{\karrow{\kDen}{(\ksum{\kDen}{\kunit})}}{
@@
 \caption{\small The initial constructor context, $\fvalCtx_0$, defines the $\fvar{Arrow}$ type constructor.
-The introductory form, $\elam{x}{\tau}{e}$, is built into the language but the elimination form goes through the operator constructor $\fvar{Arrow}.\opvar{op}$.}
+The introductory form, $\elam{x}{\tau}{e}$, is built into the language but the elimination form goes through the operator constructor $\fvar{Arrow}.\opvar{ap}$.}
 \label{arrow}
 \end{figure}
+User-defined operator constructors are declared using \textsf{opcon}.
+For reasons that we will discuss, our calculus associates every operator  constructor with a type constructor.
+The \emph{fully-qualified name} of every operator constructor, e.g.
+$\fvar{Nat}.\opvar{z}$, must be unique.
+Operator constructors, like type constructors, declare an index kind, $\kappaidx$.
+In our first example, all the operator constructors are indexed trivially (by index kind $\kunit$), but other examples use more interesting indices.
+For example, $\fvar{LabeledTuple}.\opvar{prj}$ is the operator constructor used to access a field of a labeled tuple, so it has index kind $\klabel$.
+An operator itself is, notionally, selected by indexing an operator constructor, e.g.
+$\fvar{Nat}.\opvar{s}\langle \tunit \rangle$, but technically neither operator constructors nor operators are first-class at any level (additional machinery would be needed, e.g.
+an \textsf{Op} kind, but this is not fundamental to our calculus).
+Instead, in the external language, an operator constructor is applied by simultaneously providing an index and  $n \geq 0$ \emph{arguments}, written $\eop{Tycon}{op}{\tauidx}{\splat{e}{1}{n}}$\footnote{It may be helpful to distinguish between type/operator constructors and \emph{term formers}, which are the metatheoretic entities that ultimately define the syntax.
+There are term formers at all levels in the calculus.
+For example, operator constructor application and $\lambda$ are  external term formers, and type constructor application is a type-level term former.
+An abstract syntax following Harper's conventions highlights this distinction \cite{pfpl}: the term formers $\mathtt{lam}$, $\mathtt{tcapp}$ and $\mathtt{ocapp}$, define, respectively the forms $\mathtt{lam}[\tau](x.e)$, $\texttt{tcapp}[\fvar{Tycon}](\tau)$ and $\mathtt{ocapp}[\fvar{Tycon}, \opvar{op}, \tauidx](\splat{e}{1}{n})$.}.
+For example, on line 18 of Fig.
+\ref{nat}, we see the operator constructors $\fvar{Nat}.\opvar{z}$ and $\fvar{Nat}.\opvar{s}$ being applied to compute $two$.
+In SML, the projection operator \verb|#3| can be applied to an $n$-tuple, $e$, iff $n \geq 3$.
+Note that it thus cannot be a function with a standard arrow type.
+Notionally, \verb|#| is an operator constructor and \verb|3| is its index.
+In an active embedding of $n$-tuples into @$\lambda$, this would be written $\eop{Tuple}{prj}{3}{e}$ (we will nearly recover ML's syntax later).
+%\footnote{Although our focus here is entirely on semantics, a brief note on syntax: in the expanded syntax, the trivial indices and empty argument lists can be omitted, so we could write \texttt{Nat.s(Nat.s(Nat.z))}.
+With the ability to ``open'' a type's operators into the context, we could shorten this still to \texttt{s(s(z))}.
+Alternatively, with the ability to define a TSL in a manner similar to that in Sec.
+\ref{aparsing}, we might instead just write \texttt{2}.}
+
+\subsection{Operator Definitions and Derivates}
+The expressive and metatheoretic power of the calculus arises from how the rules for the active typing judgement handle this  operator constructor application form.
+Rather than fixing the specification of a finite collection of operator constructors and tasking the \emph{compiler} with deciding a typing derivation on its basis, the specification instead tasks the \emph{operator constructor's definition}: {a type-level function that must decide the type assignment and the translation on the basis of the index and arguments, or decide that this is not possible}.
+As indicated by rule \textsc{k-opcon} in Fig.
+\ref{kindprog}, an operator constructor with index kind $\kappaidx$ must have a definition of kind $$\kappaidx \rightarrow \klist{\kDen} \rightarrow (\kDen + \kunit)$$ As this suggests, the active typing judgement  will provide the operator index and a list of {recursively determined \emph{derivates}}, which have kind $\kDen$, for the arguments and ask the definition to return a value of ``option kind'', $\ksum{\kDen}{\kunit}$, where the trivial case indicates that a derivate cannot be constructed due to an invalid index, an incorrect number of arguments or an argument of an invalid type.\footnote{In practice, we would require operator constructor providers to report information about the precise location of the error (e.g.
+which argument was of incorrect type) and provide an appropriate error message and other metadata, but we omit this in the semantics.}
+
+A \emph{derivate} is introduced in operator definitions by the form $\tden{\ibar}{\tau}$, where $\tau$ is a type assignment and has kind $\kTypeBlur$ and $\ibar$ is a \emph{translational internal term}.
+The syntax for translational internal terms mirrors that of internal terms, $\iota$, adding an additional form, $\itransof{\tauD}$, that permits one translation to be composed using the translations of other derivates (e.g.
+the derivate of an argument).
+Fig.
+\ref{tlkind} shows the kinding rules for derivates and translational internal terms (we will discuss checked derivates shortly; they are only used internally by the semantics).
+The type assignment can be extracted from a derivate using $\ttypeof{\tauD}$.
+
+Let us return to Fig. \ref{nat} to build intuition. 
+The definition of $\fvar{Nat}.\opvar{z}$ in Fig.
+\ref{nat} is quite simple: it returns the derivate $\tden{{0}}{\ttype{Nat}{\tunit}}$ if no arguments were provided, and indicates an error otherwise.
+This is done by calling a simple helper function, $\tvar{arity0}$, seen in Fig.
+\ref{helpers}.
+The definition of $\fvar{Nat}.\opvar{s}$ is only slightly more complex, because it requires inspecting a single argument.
+The helper function $\tvar{arity1}$ checks the arity of the argument list.
+If it is correct, it passes the single derivate to the shown continuation, returning the error case otherwise (which we give a more descriptive name, $\tvar{failure}$).
+The continuation we provided checks that the type of the argument is $\ttype{Nat}{\tunit}$ using a similar helper function, $\tvar{ifeq}$.
+If so, it constructs the derivate $\tden{\itransof{\tvar{d}}+1}{\ttype{Nat}{\tunit}}$ and $\tvar{ifeq}$ wraps it with $\tvar{ok}$.
+We will return to the recursor shortly.
+
+These two operator definitions make the following rules admissible in a constructor context, let us call it $\fvalCtx_{\mathbf{T}}$, containing the initial constructor context (Fig.
+\ref{arrow}) and the constructors of  Fig.
+\ref{nat} and any typing context well-kinded under that constructor context (the rules governing context kinding are in the appendix):
+\newcommand{\fvalT}{\fvalCtx_{\mathbf{T}}}
+\begin{mathpar}
+\inferrule[gt-z]{ }{
+	\ecompiles{\Gamma}{\fvalT}{\eop{Nat}{z}{\tunit}{}}{\ttype{Nat}{\tunit}}{0}
+}
+
+\inferrule[gt-s]{
+	\ecompiles{\Gamma}{\fvalT}{e}{\ttype{Nat}{\tunit}}{\iota}
+}{
+	\ecompiles{\Gamma}{\fvalT}{\eop{Nat}{s}{\tunit}{e}}{\ttype{Nat}{\tunit}}{\iota + 1}
+}
+\end{mathpar}
+
+ We will consider the complexities shortly, but it requires little more than basic intuition to see that we can strongly embed the types and introductory forms of G\"odel's $\mathbf{T}$ in @$\lambda$ quite directly (adding essentially only a barrage of trivial indices which will be removed syntactically in the expanded syntax).
+If we keep the constructor context fixed, we should be able to prove the embedding correct inductively.
+We will define this embedding later\todo{refer to it now}.
+
+ Figure \ref{att} shows how $\fvalCtx_{\mathbf{T}}$ would be built up and how these conclusions could be derived.
+The active typing judgement simply defers to the \emph{checked active typing judgement} (Fig.
+\ref{absatt}) to first produce a type assignment and a translational internal term, called the \emph{abstract translation}.
+It then \emph{deabstracts} the abstract translation to produce a type assignment and translation (Fig.
+\ref{deabstraction}).
+
+ The rule \textsc{att-op} shows how operator constructor definitions are invoked.
+First, type assignments  and translational internal terms are recursively derived.
+Then, the operator constructor definition, $\taudef$ is extracted from the constructor context and applied to the index, $\tauidx$, and a list of \emph{checked derivates} constructed from the derivations produced in the first step.
+If this produces a derivate, $\tauD$, via the left case of the sum, then it is itself, finally, \emph{checked} to produce a type assignment, $\tau$, and translational internal term, $\ibar$.
+
+The normalization rules for type-level terms are shown in Fig. \ref{tleval} and are rather uninteresting. We will discuss their metatheoretic properties in Sec. \ref{theory}.
+
+ Derivate checking, shown in Fig. \ref{ait}, is described in the next two subsections. 
+ Note that derivate checks are still static checks; they do not change the translation.
+Derivate checking is sufficient to guarantee type safety and preclude interfere issues between extensions, as we will describe, because taking a path through an operator constructor definition that produces a problematic derivate will fail.
+If providers do not want to burden clients with this variety of error, which would require them to debug the constructor definition itself, we will discuss how constructor correctness could be verified mechanically by strengthening the kind system in Sec.
+\ref{limitations}.
+%We will discuss how to prove that an extension  operator definition could be \emph{proven} to not produce such derivates (so that users only ever face type errors arising from their own code, not also extension correctness issues) in Sec.
+\ref{limitations}.
+The checks are sophisticated enough that they cannot be eliminated by a simple kind system.
+
+%$\tvar{pop\_final} : \karrow{\karrow{\klist{\kDen}}{(\karrow{\kTypeBlur}{\karrow{\kITerm}{(\ksum{\kDen}{\kunit})})}}}{(\ksum{\kDen}{\kunit})}$ ``pops'' a derivate from the head of the list and, if no other arguments remain, passes its translation and type to the ``continuation'', returning the error case otherwise.
+The continuation checks if the argument type is equal to $\ttype{Nat}{\tunit}$ using $\tvar{check\_type} : \karrow{\kTypeBlur}{\karrow{\kTypeBlur}{\karrow{\kDen}{(\ksum{\kDen}{\kunit})}}}$, which operates similarly.
+If these arity and type checks succeed, the resulting derivate is composed by adding one to the translation of the argument, passed into the continuation as $\tvar{x}$ in our example, and pairing it with the natural number type: $\tden{\titerm{\iup{\tvar{x}} + 1}}{\ttype{Nat}{\tunit}}$.
+We will return to the definition of $\fvar{Nat}.\opvar{rec}$ after in the next subsection.
+
+%We leave adding the ability to declare and manipulate new contexts as future work.
+
+\subsection{Representational Consistency Implies Type Safety}\label{repcon}
 \begin{figure}[t]
@@
 \end{mathpar}
+\caption{\small Active typing judgement for programs and external terms. The deabstraction ($\lightning$) judgement is in Fig. \ref{deabstraction}.}
+\label{att}
+%\vspace{-10pt}
+\end{figure}
+\begin{figure}[t]
+\small
 $\fbox{\inferrule{}{\ecompilesAX{e}{\tau}{\ibar}}}$
@@
 	\tEvalX{\tau_1}{\tau_1'}\\
+%	\ddbar{\fvar{Arrow}}{\fvalCtx}{\trepof{\tau_1'}}{\sbar_1}\\
 	%\delfromtau{$\Xi_0$}{\fvalCtx}{\tau_1'}{\sabs}\\\\
@@
 \inferrule[att-op]{
-	\fval{Tycon}{-}{-}{\theta} \in \fvalCtx\\
-	\tops{op}{\kappaidx}{i}{a}{\taudef} \in \theta\\
 %	\tEvalX{\tauidx}{\tau_{\text{idx}}'}\\\\
@@
 	\ecompilesA{\etCtx}{\fvalCtx}{e_n}{\tau_n}{\ibar_n}\\
-	\tEvalX{\taudef~\tauidx~(\tden{\ibar_1}{\tau_1}^\checkmark :: \cdots :: \tden{\ibar_n}{\tau_n}^\checkmark :: \tnil{\kDen})}{\tinl{\kunit}{\tden{\ibar}{\tau}}}\\
-	\ddbar{\fvar{Arrow},\fvar{Tycon}}{\fvalCtx}{\trepof{\tau}}{\sbar}\\
-	\eCtxTogCtx{\fvar{Arrow},\fvar{Tycon}}{\fvalCtx}{\etCtx}{\gtCtx}\\
-	\checkRC{\gtCtx}{\fvar{Arrow},\fvar{Tycon}}{\fvalCtx}{\ibar}{\sbar}
-}{
-	\ecompiles{\etCtx}{\fvalCtx}{\eop{Tycon}{op}{\tauidx}{\splat{e}{1}{n}}}{\tau}{\ibar}
+	\fval{Tycon}{-}{-}{\theta} \in \fvalCtx\\
+	\tops{op}{\kappaidx}{i}{a}{\taudef} \in \theta\\
+	\tEvalX{\taudef~\tauidx~(\tden{\ibar_1}{\tau_1}^\checkmark :: \cdots :: \tden{\ibar_n}{\tau_n}^\checkmark :: \tnil{\kDen})}{\tinl{\kunit}{\tauD}}\\
+	\dcheck{\Gamma}{\fvalCtx}{\fvar{Arrow}, \fvar{Tycon}}{\tauD}{\tau}{\ibar}
+}{
+	\ecompilesA{\etCtx}{\fvalCtx}{\eop{Tycon}{op}{\tauidx}{\splat{e}{1}{n}}}{\tau}{\ibar}
 }
 \end{mathpar}
-\caption{\small Active typechecking and translation.
-Normalization judgement for type-level terms is specified in Fig.
+\caption{\small The checked active typing judgement.
+The normalization judgement for type-level terms ($\Downarrow$) is specified in Fig.
 \ref{tleval}.
-Local concretization ($\leadsto$)  and abstract internal typechecking ($\sim$) judgements are specified in Fig.
-\ref{?}.
-Full concretization ($\lightning$) is specified in Fig.
-\ref{?}.}
-\label{att}
-\vspace{-10pt}
+Derivate checking is specified in Fig.
+\ref{ait}.
+Note that checked derivates can \emph{only} be constructed by the \textsc{att-op} rule (we enforce this judgementally in Sec.
+\ref{theory}).}
+\label{absatt}
 \end{figure}
+In our example, $\fvar{Nat}.\opvar{z}$ produces a translation of internal type $\dint$.
+For the translation produced by $\fvar{Nat}.\opvar{s}$ to be well-typed requires that $\itransof{\tvar{d1}}$ be of internal type $\dint$.
+Because we have checked that $\ttypeof{\tvar{d1}}$ is $\ttype{Nat}{\tunit}$ and the only other introductory form is $\fvar{Nat}.\opvar{z}$, this is not an issue.
+Well-typed external terms of type $\ttype{Nat}{\tunit}$ will always translate to well-typed internal terms of internal type $\dint$ as given.
+
+This would not be true if, for example, we added an operator constructor $\fvar{Nat}.\opvar{z2}$ that produced the derivate $$\tden{{(0, ())}}{\ttype{Nat}{\tunit}}$$
+\noindent
+In this case, there would be two different internal types, $\dint$ and $\dpair{\dint}{\dunit}$,  associated with a single external type, $\ttype{Nat}{\tunit}$.
+This makes it impossible to reason \emph{compositionally} about the translation of an external term of type $\ttype{Nat}{\tunit}$, so our implementation of $\fvar{Nat}.\opvar{s}$ would produce ill-typed translations in some cases but not others.
+Similarly, we wouldn't be able to write functions over all natural numbers because there would not be a well-typed translation to give to such a function.
+This violates type safety: there are now well-typed external terms, according to the active typing judgement, for which evaluating the corresponding translation would ``go wrong''.
+
+To reason compositionally about the semantics of well-typed external terms when they are given meaning by translation to a typed internal language, we must have that the statics maintains the following property: for every  type, $\tau$, there must exist an internal type, $\sigma$, called its \emph{representation type}, such that the translation of every external term of type $\tau$ has internal type $\sigma$.
+This principle of \emph{representational consistency} arises essentially as a strengthening of the inductive hypothesis necessary to prove that all well-typed external terms translate to well-typed internal terms, precisely because $\lambda$ and operators like $\fvar{Nat}.\opvar{s}$ are defined compositionally.
+It is closely related to the concept of \emph{type-preserving compilation} developed by Morrisett et al.
+for the TIL compiler for SML \cite{tarditi+:til-OLD}.
+%That is, instead of being used as a necessary condition for compiler correctness, we are using it as a sufficient condition for type safety.
+%We emphasize the distinction between translation (which in our calculus endows terms with a dynamic semantics) and compilation (which must preserve the dynamic semantics of terms).
+%In the presence of extensions, checking for representational consistency becomes subtle if we wish to guarantee conservativity, as we will discuss in the next subsection.
+
+For the semantics to ensure that representational consistency is maintained by all operator definitions, we require, as briefly mentioned in Sec.
+\ref{types}, that each type constructor must declare a \emph{representation schema} after the keyword \textsf{schema}.
+This must be a type-level function of kind $\karrow{\kappaidx}{\kIType}$, where $\kappaidx$ is the index kind of the type constructor.
+When the semantics (i.e.
+the compiler) needs to determine the representation type of the type $\ttype{Tycon}{\tauidx}$ it simply applies the representation schema of $\fvar{Tycon}$ to ${\tauidx}$.
+
+The kind $\kIType$ has introductory form $\titype{\bar \sigma}$ and no elimination form.
+It is not an equality kind.
+There are two forms of \emph{translational internal type}, $\bar \sigma$, that do not correspond to forms in $\sigma$ that, as with translational internal terms, allow  compositional formation:
+\begin{enumerate}
+\item $\dup{\tau}$ ``unquotes'' the translational internal type $\tau$, so that internal types can be formed programmatically\footnote{An analagous form could be added to translational internal terms, but the same effect can be achieved by working directly with derivates, albeit with some inconvenience when ``temporary intermediates'' are needed.} \todo{example}
+\item $\trepof{\tau}$ refers to the representation type of type $\tau$. This type may have come from an argument derivate or extracted from a type or operator index.
+\end{enumerate}
+
+These additional forms are not needed by the representation schema of $\fvar{Nat}$  because it is trivially indexed.
+In Fig.
+\ref{tuple}, we show an example of the type constructor $\fvar{Tuple}$, implementing the semantics of $n$-tuples by translation to nested binary products.
+Here, the representation schema requires referring to the representations of the tuple's constituent types, given in the type index\todo{add this example from appendix of ESOP}.
+
+Translational internal terms also contain translational internal types.
+We see this in the definition of the recursor on natural numbers, $\fvar{Nat}.\opvar{rec}$.
+After checking the arity, extracting the derivates of its three arguments, and checking their types (using \fvar{Arrow} to bind variables in the third position), it produces a derivate that implements the necessarily recursive dynamic semantics using a fixpoint computation in the internal language.
+This derivate's type assignment is the arbitrary type, $\tvar{t2}$.
+We cannot know what the representation type of $\tvar{t2}$ is (even in a closed constructor context, because the schema is not directly callable), so we refer to it abstractly using $\trepof{\tvar{t2}}$.
+Luckily, the translation is well-typed \emph{independent of the representation type}.
+In other words, a proof of representational consistency of the derivate produced by the definition of $\fvar{Nat}.\opvar{rec}$ is parametric (in a metamathematical sense here) over the representation type of $\tvar{t2}$.
+In fact, we will demand this for all types constructed by a type constructor other than the one an operator constructor is associated with.
+This leads us into the concept of conservativity.
+
+\subsection{Representational Independence Implies Conservativity}\label{repind}
+In isolation (let us call this language @$\lambda[\fvalCtx_{\mathbf{T}}]$) our definition of natural numbers can be shown to be a strong encoding of the semantics of G\"odel's $\mathbf{T}$.
+That is, there is a mapping of terms and types into @$\lambda[\fvalCtx_{\mathbf{T}}]$ and an inverse mapping.
+Indeed, these are the entirely obvious one, given how directly we constructed @$\lambda[\fvalCtx_{\mathbf{T}}]$.
+
+Typing is preserved by  these mappings, and we can show by a bisimulation argument that the translation produced by @$\lambda[\fvalCtx_{\mathbf{T}}]$ implements the dynamic semantics of G\"odel's $\mathbf{T}$.
+We must leave full details to an appendix, but the techniques are fundamentally standard when reasoning about compiler correctness, here lifted into the language.
+A necessary lemma in the proofs, however, is that the value of every translation of an external term of type $\ttype{Nat}{\tunit}$ is a non-negative integer.
+This is needed to show that the fixpoint computation in the definition of $\fvar{Nat}.\opvar{rec}$ is terminating, as the recursor always does in $\mathbf{T}$.
+A very similar lemma is needed to show that the translation of every external term of type $\ttype{Nat}{\tunit}$ is equivalent to the translation that would be produced by some combination of applications of $\fvar{Nat}.\opvar{z}$ and $\fvar{Nat}.\opvar{s}$ (corresponding to the canonical forms lemma of $\mathbf{T}$).
+
+Fortunately, the definitions we have provided thus far admit this lemma.
+Indeed, the proof is quite straightforward.
+However, it is also quite precarious because it relies on exhaustive induction over the available operators.
+As soon as we load another operator or type constructor into the language, these theorems may no longer be \emph{conserved}.
+We already saw an example where a representationally inconsistent operator caused problems, but the semantics render uses of it ill-typed using the representation schema (in a manner we will make more precise shortly).
+But there are operator constructors that are representationally consistent, but still cause problems because they violate internal invariants.
+For example, $$\tops{n1}{\kunit}{i}{a}{\tden{-1}{\ttype{Nat}{\tunit}}}$$
+
+Were this operator constructor introduced inside $\fvar{Nat}$, there is no way for our semantics to distinguish between an incorrect implementation of $\fvar{Nat}$ and a correct implementation of some other type where non-termination of the recursor was the intended behavior (without a separate specification).
+
+But in any \emph{other} type constructor, this clearly appears to violate the tacit, thus far, abstraction barrier between type constructors.
+In particular, if we extended @$\lambda[\fvalCtx_{\mathbf{T}}]$ with such a type constructor, let us call it $\fvar{Evil}$, then our crucial lemma relating to non-negativity would not be conserved, and thus, our carefully constructed strong encoding would no longer be a strong encoding.
+In particular, our canonical forms lemma now has a new case, and the recursor behaves badly.
+\emph{If we assume an abstraction barrier between type constructors\footnote{A fragment consisting of finitely many type constructors can be easily turned into just one by adding a sum type to the index and case analyzing against it whenever a distinction is necessary.} must exist, we do not need need a specification to see that $\fvar{Evil}.\opvar{n1}$ is violating it}.
+We can prevent problems by using a technique closely related to the one used to enforce abstraction barriers between modules -- holding the representation type abstract.
+If $\fvar{Evil}.\opvar{n1}$ cannot know that the representation type of $\ttype{Nat}{\tunit}$ is $\dint$, then its definition is incorrect independent of the specification of $\fvar{Nat}$.
+\section{Metatheory}\label{theory}
+\subsection{Active Typing}
 \begin{figure}[t]
 \small
-$\fbox{\inferrule{}{\ddbarX{\sbar}{\sbar'}}}$
+\noindent
+$\fbox{\inferrule{}{\dcheckX{\tauD}{\tau}{\ibar}}}$
+~~~$\Xi ::= \emptyset \pipe \Xi, \fvar{Tycon}$
+\begin{mathpar}
+\inferrule[d-check]{
+	\ddbar{\Xi}{\fvalCtx}{\trepof{\tau}}{\sabs}\\
+%	\eCtxTogCtx{\Xi}{\fvalCtx}{\etCtx}{\gtCtx}\\
+	\checkRC{\emptyctx}{\Xi}{\fvalCtx}{\ibar}{\sabs}
+}{
+	\dcheckX{\tden{\ibar}{\tau}}{\tau}{\ibar}
+}
+
+\inferrule[d-checked]{
+	%\dcheckX{\tden{\ibar}{\tau}}{\tau}{\ibar}
+}{
+	\dcheckX{\tden{\ibar}{\tau}^\checkmark}{\tau}{\ibar}
+}
+\end{mathpar}
+$\fbox{\inferrule{}{\ddbarX{\sbar}{\sabs}}}$
+~~~$\sabs ::= \sabs \rightharpoonup \sabs' \pipe \cdots \pipe \sabsrep{\tau}$
 %~~~$\fvalCtx ::= \fvalCtx_0 \pipe \fvalCtx, \fvalDf$~~~
-~~~$\Xi ::= \emptyset \pipe \Xi, \fvar{Tycon}$~~~
-%~~~$\Xi_0 := \fvar{Arrow}$
 \begin{mathpar}
@@
 %}
-\inferrule[local-conc-show-rep]{
+\inferrule[abs-trity-parr]{
+	\ddbarX{\sbar_1}{\sabs_1}\\
+	\ddbarX{\sbar_2}{\sabs_2}
+}{
+	\ddbarX{\darrow{\sbar_1}{\sbar_2}}{\darrow{\sabs_1}{\sabs_2}}
+}
+
+\inferrule[abs-trity-cancel]{
+	\ddbarX{\sbar}{\sabs}
+}{
+	\ddbarX{\dup{\titype{\sbar}}}{\sabs}
+}
+
+\inferrule[abs-trity-concrete]{
 	\fvar{Tycon} \in \Xi\\
@@
 	\tEvalX{\taurep~\tauidx}{\titype{\sbar}}\\
-	\ddbarX{\sbar}{\sbar'}
-}{
-	\ddbarX{\trepof{\ttype{Tycon}{\tauidx}}}{\sbar'}
-}
-
-\inferrule[local-conc-hide-rep]{
+	\ddbarX{\sbar}{\sabs}
+}{
+	\ddbar{\Xi}{\fvalCtx}{\trepof{\ttype{Tycon}{\tauidx}}}{\sabs}
+}
+
+\inferrule[abs-trity-abstract]{
 	\fvar{Tycon} \notin \Xi
 }{
-	\ddbarX{\trepof{\ttype{Tycon}{\tauidx}}}{\trepof{\ttype{Tycon}{\tauidx}}}
-}
-
-\inferrule[local-conc-unquote]{
-	\ddbarX{\sbar}{\sbar'}
-}{
-	\ddbarX{\dup{\titype{\sbar}}}{\sbar'}
-}
-
-%
-\inferrule[local-conc-parr]{
-	\ddbarX{\sbar_1}{\sbar_1'}\\
-	\ddbarX{\sbar_2}{\sbar_2'}
-}{
-	\ddbarX{\darrow{\sbar_1}{\sbar_2}}{\darrow{\sbar_1'}{\sbar_2'}}
+	\ddbarX{\trepof{\ttype{Tycon}{\tauidx}}}{\sabsrep{\ttype{Tycon}{\tauidx}}}
 }
@@
 
-\text{\color{gray} (remaining rules follow recursively, omitted)}
+\text{\color{gray} (omitted rules follow recursively like \textsc{abs-it-parr})}
 \end{mathpar}
-$\fbox{\inferrule{}{\eCtxTogCtxX{\etCtx}{\gtCtx}}}$
-~~~$\gtCtx ::= \emptyctx \pipe \gtCtxX{x}{\sbar}$
-\begin{mathpar}
-\inferrule[abs-empty]{ }{
-	\eCtxTogCtxX{\emptyctx}{\emptyctx}
-}
-
-\inferrule[abs-ctx]{
-	\eCtxTogCtxX{\etCtx}{\gtCtx}\\
-	\ddbarX{\trepof{\tau}}{\sbar}
-}{
-	\eCtxTogCtxX{\etCtxX{x}{\tau}}{\gtCtxX{x}{\sbar}}
-}
-\end{mathpar}
-$\fbox{\inferrule{}{\checkRCX{\ibar}{\sbar}}}$
+%$\fbox{\inferrule{}{\eCtxTogCtxX{\etCtx}{\gtCtx}}}$
+%~~~$\gtCtx ::= \emptyctx \pipe \gtCtxX{x}{\sbar}$
+%\begin{mathpar}
+%\inferrule[abs-empty]{ }{
+%	\eCtxTogCtxX{\emptyctx}{\emptyctx}
+%}
+%
+%\inferrule[abs-ctx]{
+%	\eCtxTogCtxX{\etCtx}{\gtCtx}\\
+%	\ddbarX{\trepof{\tau}}{\sbar}
+%}{
+%	\eCtxTogCtxX{\etCtxX{x}{\tau}}{\gtCtxX{x}{\sbar}}
+%}
+%\end{mathpar}
+$\fbox{\inferrule{}{\checkRCX{\ibar}{\sabs}}}$
+~~~$\gtCtx ::= \emptyset \pipe \gtCtx, x \sim \sbar$
 \begin{mathpar}
@@
 \inferrule[abs-i-fix]{
-	\ddbarX{\sbar}{\sbar'}\\
-	\checkRC{\gtCtxX{x}{\sbar'}}{\Xi}{\fvalCtx}{\ibar}{\sbar'}
-}{
-	\checkRCX{\ifix{x}{\sbar}{\gabs}}{\sbar'}
+	\ddbarX{\sbar}{\sabs}\\
+	\checkRC{\gtCtxX{x}{\sabs}}{\Xi}{\fvalCtx}{\ibar}{\sabs}
+}{
+	\checkRCX{\ifix{x}{\sbar}{\ibar}}{\sbar'}
 }
@@
 
-\text{\color{gray} (remaining rules follow from a standard statics  similarly, omitted)}
+\text{\color{gray} (omitted rules follow from a standard statics  similarly)}
 
@@
 %
-\inferrule[show-trans]{
+\inferrule[abs-trans-local-unchecked]{
 	\fvar{Tycon} \in \Xi\\
-	\checkRCX{\ibar}{\sbar}
-}{
-	\checkRCX{\itransof{\tden{\ibar}{\ttype{Tycon}{\tauidx}}}}{\sbar}
-}
-
-\inferrule[hide-trans]{
+	\ddbarX{\trepof{\ttype{Tycon}{\tauidx}}}{\sabs}\\
+	\checkRCX{\ibar}{\sabs}
+}{
+	\checkRCX{\itransof{\tden{\ibar}{\ttype{Tycon}{\tauidx}}}}{\sabs}
+}
+
+\inferrule[abs-trans-local-checked]{
+	\fvar{Tycon} \in \Xi\\
+	\ddbarX{\trepof{\ttype{Tycon}{\tauidx}}}{\sabs}
+}{
+	\checkRCX{\itransof{\tden{\ibar}{\ttype{Tycon}{\tauidx}}^\checkmark}}{\sabs}
+}
+
+\inferrule[abs-trans-foreign-checked]{
 	\fvar{Tycon} \notin \Xi
 }{
-	\checkRCX{\itransof{\tden{\ibar}{\ttype{Tycon}{\tauidx}}^{\checkmark}}}{\trepof{\ttype{Tycon}{\tauidx}}}
+	\checkRCX{\itransof{\tden{\ibar}{\ttype{Tycon}{\tauidx}}^{\checkmark}}}{\sabsrep{\ttype{Tycon}{\tauidx}}}
 }
 \end{mathpar}
-\caption{\small Abstracted internal typing}
+\caption{\small Abstracted internal typing involves typechecking the internal language while holding the representation of all types other than those in the current fragment, listed in the fragment context, $\Xi$, abstract.
+They key rule is \textsc{abs-trans-foreign-checked}, permitting the translation of a checked derivate that has a type assignment outside the fragment to be used against that type's abstract representation type.}
 \label{ait}
 \end{figure}
-User-defined operator constructors are declared using \textsf{opcon} (cf.
-Fig.
-\ref{kindprog}).
-For reasons that we will discuss, our calculus associates every operator  constructor with a type constructor.
-The \emph{fully-qualified name} of every operator constructor, e.g.
-$\fvar{Nat}.\opvar{z}$, must be unique.
-Operator constructors are indexed by type-level values, like type constructors, and so also declare an index kind, $\kappaidx$.
-In our first example, all the operator constructors are indexed trivially (by index kind $\kunit$), but later examples will use more interesting indices.
-For example, the projection operator constructor for labeled tuples, $\fvar{LabeledTuple}.\opvar{prj}$, has index kind $\klabel$.
-An operator itself is, notionally, selected by indexing an operator constructor, e.g.
-$\fvar{Nat}.\opvar{s}\langle \tunit \rangle$, but note that neither operator constructors nor operators are first-class at any level (additional machinery would be needed, and this is not fundamental to our calculus).
-
-Instead, in the external language, an operator constructor is applied by simultaneously selecting an operator and invoking it on $n \geq 0$ \emph{arguments}, written $\eop{Tycon}{op}{\tauidx}{\splat{e}{1}{n}}$\footnote{It may be helpful to distinguish between type/operator constructors and \emph{term formers}, which are the metatheoretic entities that ultimately define the syntax.
-There are term formers at all levels in the calculus.
-For example, operator constructor application and $\lambda$ are  external term formers, and type constructor application is a type-level term former.
-An abstract syntax following Harper's conventions highlights this distinction \cite{pfpl}: the term formers $\mathtt{lam}$, $\mathtt{tcapp}$ and $\mathtt{ocapp}$, define, respectively the forms $\mathtt{lam}[\tau](x.e)$, $\texttt{tcapp}[\fvar{Tycon}](\tau)$ and $\mathtt{ocapp}[\fvar{Tycon}, \opvar{op}, \tauidx](\splat{e}{1}{n})$.}.
-For example, on line 18 of Fig.
-\ref{nat}, we see the operator constructors $\fvar{Nat}.\opvar{z}$ and $\fvar{Nat}.\opvar{s}$ being applied to compute $two$.
-%\footnote{Although our focus here is entirely on semantics, a brief note on syntax: in the expanded syntax, the trivial indices and empty argument lists can be omitted, so we could write \texttt{Nat.s(Nat.s(Nat.z))}.
-With the ability to ``open'' a type's operators into the context, we could shorten this still to \texttt{s(s(z))}.
-Alternatively, with the ability to define a TSL in a manner similar to that in Sec.
-\ref{aparsing}, we might instead just write \texttt{2}.}
-To derive the active typing judgement for an operator constructor invocation, the semantics calls   
- into the operator constructor's \emph{definition}: a type-level function that must decide a type assignment and a translation on the basis of the index and arguments, or decide that this is not possible.
-More specifically, in @$\lambda$, examine the provided operator index and the recursively determined \emph{derivates} of the arguments to determine a derivate for the operation as a whole,  or indicate an error.
-A derivate is notionally a type-level representation of the result of deriving the active typing judgement described\footnote{Technically, the abstracted active typing judgement, introduced shortly.}: an internal term, called the {translation},  paired with a type.
-Derivates have kind $\kDen$ and introductory form $\tden{\tautrans}{\tautype}$.
-The elimination form $\tderbind{x}{t}{\tau}{\tau'}$ extracts the translation and type from the derivate $\tau$, binding it to $\tvar{x}$ and $\tvar{t}$ respectively in $\tau'$.
-%Indeed, one might be tempted to simply define derivates as products.
-However, we treat translations with much care throughout the calculus, so this is not quite the case (we will return to this shortly).
-Because constructing a derivate is not always possible (e.g.
-when there is a type error in the client's code, or an invalid index was provided), the return kind of the function is an ``option kind'', $\ksum{\kDen}{\kunit}$, where the trivial case indicates a statically-detected error in the client's program.
-In practice, it would instead require providers to report information about the precise location of the error and an appropriate error message.
-
-A translation, as we have said, is an internal term.
-To construct an internal term using a type-level function, as we must do to construct a derivate, we must expose a type-level representation of the internal language.
-A \emph{quoted internal term} is a type-level value of kind $\kITerm$ with introductory form $\titerm{\bar \iota}$ and a \emph{quoted internal type} is a type-level value of kind $\kIType$ with introductory form $\titype{\bar \sigma}$.
-Neither kind has an elimination form.
-Instead, the syntax for the quoted  internal language includes complementary \emph{unquote forms} $\iup{\tau}$ and $\dup{\tau}$ that permit the interpolation of another quoted term or type, respectively, into the one being formed.
-Interpolation is capture-avoiding, as our semantics will clarify.
-We have now described all the kinds in our language.
-
-The definition of $\fvar{Nat}.\opvar{z}$ is quite simple: it returns the derivate $\tden{\titerm{0}}{\ttype{Nat}{\tunit}}$ if no arguments were provided, and indicates an error (an \emph{arity error}, though we do not distinguish this in our calculus) otherwise.
-This is done by calling a simple helper function, $\tvar{is\_empty} : \karrow{\klist{\kDen}}{\karrow{\kDen}{(\ksum{\kDen}{\kunit})}}$, that selects the appropriate case of the sum given the derivate that should be produced if the list is indeed empty.
-The definition of $\fvar{Nat}.\opvar{s}$ is only slightly more complex, because it requires inspecting a single argument.
-The helper function $\tvar{pop\_final} : \karrow{\karrow{\klist{\kDen}}{(\karrow{\kTypeBlur}{\karrow{\kITerm}{(\ksum{\kDen}{\kunit})})}}}{(\ksum{\kDen}{\kunit})}$ ``pops'' a derivate from the head of the list and, if no other arguments remain, passes its translation and type to the ``continuation'', returning the error case otherwise.
-The continuation checks if the argument type is equal to $\ttype{Nat}{\tunit}$ using $\tvar{check\_type} : \karrow{\kTypeBlur}{\karrow{\kTypeBlur}{\karrow{\kDen}{(\ksum{\kDen}{\kunit})}}}$, which operates similarly.
-If these arity and type checks succeed, the resulting derivate is composed by adding one to the translation of the argument, passed into the continuation as $\tvar{x}$ in our example, and pairing it with the natural number type: $\tden{\titerm{\iup{\tvar{x}} + 1}}{\ttype{Nat}{\tunit}}$.
-We will return to the definition of $\fvar{Nat}.\opvar{rec}$ after in the next subsection.
-
-By writing the typechecking and translation logic in this way, as a total function, we are taking a rather ``implementation-focused view'' rather than attempting to extract such a function from a declarative specification.
-That is,  we leave to the provider (or to a program generator or kind-specific language that transforms such a specification into an operator definition) the problem of finding a deterministic algorithm that adequately implements their intended semantics, allowing us to prove decidability of typechecking for the language as a whole by essentially just citing the termination theorem for the type-level language.
-We also avoid difficulties with error reporting in practice.
-Rob Simmons' undergraduate thesis has a good discussion of these issues \cite{rjs-princeton}.
-
-Because the input to an operator definition is the recursively determined derivate of the argument in the same context as the operator appears in, our mechanism does not presently permit the definition of operator constructors that bind variables themselves or require a different form of typing judgement (e.g.
-additional forms of contexts).
-This is also why the $\lambda$ operator constructor needs to be built in.
-It is the only operator in our language that can be used to bind variables.
-The type constructor $\fvar{Arrow}$ can be defined from within the language (and is included in the ``prelude'' constructor context), but only defines the operator constructor, $\opvar{ap}$.
-These two operators translate to their corresponding forms in the internal language directly, as we will discuss below.
-We leave adding the ability to declare and manipulate new contexts as future work.
-
-\subsection{Representational Consistency Implies Type Safety}\label{repcon}
-
-In our example, natural numbers are represented internally as integers.
-Were this not the case -- if, for example, we added an operator constructor $\fvar{Nat}.\opvar{z2}$ that produced the derivate $\tden{\titerm{\iinl{\dint}{\iunit}}}{\ttype{Nat}{\tunit}}$, then there would be two different internal types, $\dint$ and $\dsum{\dunit}{\dint}$,  associated with a single external type, $\ttype{Nat}{\tunit}$.
-This makes it impossible to operate compositionally on the translation of an external term of type $\ttype{Nat}{\tunit}$, so our implementation of $\fvar{Nat}.\opvar{s}$ would produce ill-typed translations in some cases but not others.
-Similarly, we wouldn't be able to write functions over all natural numbers because there would not be a well-typed translation to give to such a function.
-
-To reason compositionally about the semantics of well-typed external terms when they are given meaning by translation to a typed internal language, we must have the following property: for every  type, $\tau$, there must exist an internal type, $\sigma$, called its \emph{representation type}, such that the translation of every external term of type $\tau$ has internal type $\sigma$.
-This principle of \emph{representational consistency} arises essentially as a strengthening of the inductive hypothesis necessary to prove that all well-typed external terms translate to well-typed internal terms, precisely because operators like $\lambda$ and $\fvar{Nat}.\opvar{s}$ are defined compositionally.
-It is closely related to the concept of \emph{type-preserving compilation} developed by Morrisett et al.
-for the TIL compiler for Standard ML \cite{TIL}\todo{cite}.
-%That is, instead of being used as a necessary condition for compiler correctness, we are using it as a sufficient condition for type safety.
-%We emphasize the distinction between translation (which in our calculus endows terms with a dynamic semantics) and compilation (which must preserve the dynamic semantics of terms).
-%In the presence of extensions, checking for representational consistency becomes subtle if we wish to guarantee conservativity, as we will discuss in the next subsection.
-
-It is easy to show by induction that, under a ``closed-world assumption'' where the only available operators are $\fvar{Nat}.\opvar{z}$ and $\fvar{Nat}.\opvar{s}$, the representation type of $\ttype{Nat}{\tunit}$ is $\dint$.
-If we can maintain this under an ``open-world assumption'' (requiring that, for example, applications of operator constructors like $\fvar{Nat}.\opvar{z2}$, above, are not well-typed), and we target a type safe internal language, then we will achieve type safety: well-typed external terms cannot go wrong, because they always translate to well-typed internal terms, which cannot go wrong.
-For the semantics to ensure that representational consistency is maintained by all operator definitions, we require that each type constructor must declare a \emph{representation schema} with the keyword \textsf{schema}.
-This must be a type-level function of kind $\karrow{\kappaidx}{\kIType}$, where $\kappaidx$ is the index kind of the type constructor.
-When the compiler needs to determine the representation type of the type $\ttype{Tycon}{\tauidx}$ it simply applies the representation schema of $\fvar{Tycon}$ to ${\tauidx}$.
-
-As described above, the kind $\kIType$ has introductory form $\titype{\bar \sigma}$ and no elimination form.
-There are two forms in $\bar \sigma$ that do not correspond to forms in $\sigma$ that allow quoted internal types to be formed compositionally:
-\begin{enumerate}
-\item $\dup{\tau}$, already described, unquotes (or ``interpolates'') the quoted internal type $\tau$
-\item $\trepof{\tau}$ refers to the representation type of type $\tau$ (we will see in the next subsection why this needs to be in the syntax for $\bar \sigma$ and not directly in the type-level language)
-\end{enumerate}
-
-These additional forms are not needed by the representation schema of $\fvar{Nat}$  because it is trivially indexed.
-In Fig.
-\ref{tuple}, we show an example of the type constructor $\fvar{Tuple}$, implementing the semantics of $n$-tuples by translation to nested binary products.
-Here, the representation schema requires referring to the representations of the tuple's constituent types, given in the type index\todo{add this example from appendix of ESOP}.
-
-Operator constructor definitions might also need to refer to the representation of a type.
-We see this in the definition of the recursor on natural numbers, $\fvar{Nat}.\opvar{rec}$.
-After checking the arity and extracting the types and translations of its three arguments, it produces a derivate that implements the necessarily recursive dynamic semantics using a fixpoint computation in the internal language.
-This fixpoint computation produces a result of some arbitrary type, $\tvar{t2}$.
-We cannot know what the representation type of $\tvar{t2}$ is, we refer to it abstractly using $\trepof{\tvar{t2}}$.
-The translation is well-typed no matter what the representation type is.
-In other words, a proof of representational consistency of the derivate produced by the definition of $\fvar{Nat}.\opvar{rec}$ is parametric (in the metamathematics) over the representation type of $\tvar{t2}$.
-This leads us into the concept of conservativity.
-
-\subsection{Representational Independence Implies Conservativity}\label{repind}
-In isolation, our definition of natural numbers can be shown to be a strong encoding of the semantics of G\"odel's $\mathbf{T}$.
-That is, there is a bijection between the terms and types of the two languages and the typing judgements preserve this mapping.
-Moreover, we can show by a bisimulation argument that the translation produced by @$\lambda$ implements the dynamic semantics of G\"odel's $\mathbf{T}$.
-We will provide more details on this later.
-A necessary lemma in the proof, however, is that the value of every translation of an external term of type $\ttype{Nat}{\tunit}$ is a non-negative integer.
-This is needed to show that the fixpoint computation in the definition of $\fvar{Nat}.\opvar{rec}$ is terminating, as the recursor always does in $\mathbf{T}$.
-A very similar lemma is needed to show that the translation of every external term of type $\ttype{Nat}{\tunit}$ is equivalent to the translation that would be produced by some combination of applications of $\fvar{Nat}.\opvar{z}$ and $\fvar{Nat}.\opvar{s}$ (the analog to a canonical forms lemma in our setting).
-Fortunately, the definitions we have provided thus far admit such lemmas.
-
-However, these theorems are all quite precarious because they rely on exhaustive induction over the available operators.
-As soon as we load another library into the program, these theorems may no longer be \emph{conserved}.
-For example, the following operator declaration in some other type that we have loaded would topple our house of cards\todo{TBC}:
-
-\section{Metatheory}\label{theory}
+\begin{figure}
+\small
+$\fbox{\inferrule{}{\eraseX{\sbar}{\sigma}}}$
+\begin{mathpar}
+\inferrule[deabs-it-repof]{
+	\fval{Tycon}{-}{\taurep}{-} \in \fvalCtx\\
+	\tEvalX{\taurep~\tauidx}{\titype{\sbar}}\\
+	\eraseX{\sbar}{\sigma}
+}{
+	\eraseX{\trepof{\ttype{Tycon}{\tauidx}}}{\sigma}
+}
+
+\inferrule[deabs-it-cancel]{
+	\eraseX{\sbar}{\sigma}
+}{
+	\eraseX{\dup{\titype{\sbar}}}{\sigma}
+}
+
+\inferrule[deabs-it-parr]{
+	\eraseX{\sbar_1}{\sigma_1}\\
+	\eraseX{\sbar_2}{\sigma_2}
+}{
+	\eraseX{\darrow{\sbar_1}{\sbar_2}}{\darrow{\sigma_1}{\sigma_2}}
+}
+
+\text{\color{gray} (remaining forms follow recursively like \textsc{deabs-it-parr}, omitted)}
+\end{mathpar}
+$\fbox{\inferrule{}{\eraseX{\ibar}{\iota}}}$
+\begin{mathpar}
+\inferrule[deabs-i-var]{ }{\eraseX{x}{x}}
+
+\inferrule[deabs-i-fix]{
+	\eraseX{\sbar}{\sigma}\\
+	\eraseX{\ibar}{\iota}
+}{
+	\eraseX{\ifix{x}{\sbar}{\ibar}}{\ifix{x}{\sigma}{\iota}}
+}
+
+\text{\color{gray} (remaining forms follow recursively like \textsc{deabs-i-fix}, omitted)}
+
+\inferrule[deabs-trans-unchecked]{
+	\eraseX{\ibar}{\iota}
+}{
+	\eraseX{\itransof{\tden{\ibar}{\tau}}}{\iota}
+}
+
+\inferrule[deabs-trans-checked]{
+	\eraseX{\ibar}{\iota}
+}{
+	\eraseX{\itransof{\tden{\ibar}{\tau}^\checkmark}}{\iota}
+}
+\end{mathpar}
+\caption{\small Deabstraction simply recursively eliminates syntactic abstraction barriers and composition-related constructs, and is the final step.}
+\label{deabstraction}
+\end{figure}
+
+allowing us to prove decidability of typechecking for the language as a whole by essentially just citing the termination theorem for the type-level language and proving that the underlying semantics are otherwise well-founded.
 
@@
 \section{Examples}\label{examples}
-
-\section{Limitations}\label{limitations}
+\begin{figure}
+\small
+\begin{flalign}
+\small
+ & \family{Tuple}{\klist{\kTypeBlur}}{\\
+ & \quad \tops{new}{\kunit}{\_}{a}{\tlam{\_}{\kunit}{\tlam{a}{\klist{\kDen}}{\\
+ & \quad\quad
+ 	\tfold{\tvar{a}}{\tden{{()}}{\ttype{Tuple}{\tnil{\kTypeBlur}}}}{d}{b}{r}{\\
+			& \quad\quad\quad \tfamcase{\ttypeof{\tvar{r}}}{Ntuple}{i}{\\
+			& \quad\quad\quad\quad \tfold{\tvar{b}}{\tden{\itransof{\tvar{d}}}{\ttype{Ntuple}{\tcons{\ttypeof{\tvar{d}}}{\tvar{i}}}}\\& \quad\quad\quad\quad}{\_}{\_}{\_}{\tden{\\& \quad\quad\quad\quad\quad {(\itransof{\tvar{d}},\itransof{\tvar{r}})}}{\ttype{Ntuple}{\tcons{{\ttypeof{\tvar{d}}}}{\tvar{i}}}}}
+			}{\terr}
+	}
+ }}};\\
+ & \quad \tops{prj}{\kint}{i}{a}{\lambda \tvar{i}:\kint.\lambda \tvar{a}:\klist{\kDen}. 
+ 	\tapp{\tvar{arity1}}{\tapp{\tvar{a}}{\tlam{d}{\kDen}{\\
+	& \quad\quad\tfamcase{\ttypeof{\tvar{d}}}{Ntuple}{nl}{\\
+ & \quad\quad \tfold{\tvar{nl}}{\terr}{t1}{j}{\_}{\\
+ & \quad\quad \tfold{\tvar{j}}{\tifeq{\tvar{i}}{1}{\dint}{\tden{\itransof{d}}{\tvar{t1}}}{\terr}\\&\quad\quad\quad}{\_}{\_}{\_}{\\
+ & \quad\quad\quad (\tlam{p}{\kpair{\kDen}{\dint}}{\tifeq{\tsnd{\tvar{p}}}{\tvar{i}}{\dint}{\tfst{\tvar{p}}}{\terr}})\\
+ & \quad\quad\quad (\tapp{\tvar{foldl}}{\tapp{\tvar{nl}}{\tapp{(\tden{\tvar{x}}{\tvar{nt}},0)}{
+ 	\tlam{r}{\kpair{\kDen}{\dint}}{\tlam{t}{\kTypeBlur}{\tlam{ts}{\klist{\kTypeBlur}}{\\
+	& \quad\quad\quad\quad \tifeq{\tvar{i}}{\tsnd{\tvar{r}}}{\dint}{\tvar{r}}{\tdencase{\tfst{\tvar{r}}}{rx}{\_}{\\
+	& \quad\quad\quad\quad\quad \tifeq{\tvar{i}}{\tsnd{\tvar{r}}+1}{\dint}{\\
+	& \quad\quad\quad\quad\quad\quad \tifeq{\tvar{ts}}{\tnil{\kTypeBlur}}{\klist{\kTypeBlur}}{(\tden{\tvar{rx}}{\tvar{t}},\tvar{i})\\
+	& \quad\quad\quad\quad\quad\quad}{(\tden{\titerm{\ifst{\iup{\tvar{rx}}}}}{\tvar{t}},\tvar{i})}\\	
+	& \quad\quad\quad\quad\quad}{(\tden{\titerm{\isnd{\iup{\tvar{rx}}}}}{\tvar{t}},\tsnd{\tvar{r}}+1)}}{\terr}}}}}
+ }}})
+ }}{\terr}}}}}
+ }\\ & 
+%& \quad \tops{pr}a}{b}{c}{d}
+ }{d}{i}{(\tlam{i}{\klist{\kTypeBlur}}{
+ 	\\& \quad \quad \tfold{\tvar{i}}{\titype{\dunit}}{s}{j}{r}{
+ 		\tfold{\tvar{j}}{\titype{\trepof{\tvar{s}}}}{\_}{\_}{\_}{\\
+		&\quad\quad\quad
+ 		\titype{\dpair{\trepof{\tvar{s}}}{\dup{\tvar{r}}}}
+ 		}
+ 	}
+ })}
+\end{flalign}
+\caption{ABC}
+\label{tuple}
+\end{figure}
+
+\section{Limitations and Future Work}\label{limitations}
 \todo{integrate this}Indeed, it is not a stretch to imagine a reasonably faithful implementation of our calculus starting with a pure, total subset of the term language of ML to construct the type-level language.
 We leave the development of such an implementation (and of bootstrapping type-level and internal languages that are themselves user-defined or extensible) as areas for future work to focus sharply on the theoretical foundations in this work.
+
+Because the input to an operator definition is a list of the recursively determined derivates of the argument in the same typing context as it appears in, our mechanism does not presently permit the definition of operator constructors that bind variables themselves or require a different form of typing judgement (e.g.
+additional forms of contexts).
+This is why the $\lambda$ form needs to be built in.
+It is the only form in the external language that can be used to bind variables.
+The type constructor $\fvar{Arrow}$ can, however, be defined from within the language and is included in the initial constructor context (it could be written out and included as a ``prelude'' as well).
+As seen in Fig.
+\ref{arrow}, it only defines the operator constructor $\opvar{ap}$.
+
+By writing the type assignment and translation logic in this way, as a type-level function, we are taking a rather ``implementation-focused view'' rather than attempting to extract a compiler from a declarative specification of a type system and an operational semantics.
+That is, @$\lambda$ leaves ultimately to the provider the problem of finding a decidable type assignment algorithm and resolve non-determinism that may arise.
+Similarly, we do not attempt to turn an operational semantics into a compiler automatically.
+We  also avoid difficulties with useful error reporting in practice.
+Simmons' discusses the difficulties of these issues \cite{rjs-princeton}.
+Composing two dialect specifications directly is more fraught with metatheoretic issues than analyzing two functional programs used together.
+By looking at the problem as a functional programmer, we can apply intuitions about abstraction barriers more directly.
+A solution that permits implementations to be extracted from fragment specifications could easily be layered atop this mechanism and inherit these guarantees.
+Alternatively, extension correctness could be mechanized by encoding these semantics in a proof assistant like Coq and then using its extraction mechanisms to produce a compiler.
 
@@
 \end{mathpar}
-\caption{\small Evaluation semantics for type-level terms}
+\caption{\small Normalization semantics for type-level terms, including those inside translational IL. There is nothing clever happening here.}
 \label{tleval}

### 2014-03-15 globaldsl13/motivation.tex
removed stuff

@@
   \begin{lstlisting}
-serve(page, loc) where 
-  val page = ~ 
+serve : (HTML, URL) -> unit
+serve(~, `products.nameless.com`) 
   html:
@@
           {FeedBox(servlet.hotProds())}
-  val loc = ~
-    products.nameless.com
   \end{lstlisting}

### 2014-03-15 oopsla13/att-oopsla13.tex
removed stuff

@@
 \documentclass[10pt]{sigplanconf}
+
 % The following \documentclass options may be useful:
@@
 % authoryear    To obtain author/year citation style instead of numeric.
-\usepackage{listings}
 \usepackage[usenames,dvipsnames]{color}
-\usepackage[usenames,dvipsnames]{xcolor}
+\usepackage[usenames,dvipsnames,svgnames,table]{xcolor}
 \usepackage{mathpartir}
@@
 \usepackage{graphicx}
-
-\usepackage[noindentafter]{titlesec}
+\usepackage{url}
+\newcommand{\TODO}[1]{{\color{red} #1}}
+%\usepackage{float}
+%\floatstyle{ruled}
+%\newfloat{figure}{tp}{lop}
+%\floatname{figure}{Listing}
+
+%\usepackage[noindentafter]{titlesec}
 %\titlespacing{\section}{0pt}{*0.5}{*0.25}
@@
 \input{macros-lamalpha}
+\newcommand{\ATT}{AT\&T}
 \newtheorem{theorem}{Theorem}%[document]
@@
 
+\usepackage{alltt}
+\renewcommand{\ttdefault}{txtt}
+\usepackage{color}
+\usepackage[usenames,dvipsnames,svgnames,table]{xcolor}
+\definecolor{mauve}{rgb}{0.58,0,0.82}
+\usepackage{listings}
+\lstset{
+  language=Python,
+  showstringspaces=false,
+  formfeed=\newpage,
+  tabsize=4,
+  commentstyle=\itshape,
+  basicstyle=\ttfamily\footnotesize,
+  morekeywords={lambda, self, assert, as},
+  numbers=left,
+  numberstyle=\footnotesize\color{gray}\textsf,
+  xleftmargin=2em,
+  stringstyle=\color{mauve}
+}
+\lstdefinestyle{Bash}{
+    language={}, 
+    moredelim=**[is][\color{blue}\bf\ttfamily]{`}{`},
+}
+\lstdefinestyle{OpenCL}{
+	language=C++,
+	morekeywords={kernel, __kernel, global, __global, size_t, get_global_id, sin}
+}
+
 \begin{document}
-
-%\conferenceinfo{POPL '13}{somewhere} 
-%\copyrightyear{2013} 
-%\copyrightdata{[to be supplied]}
-
-%\titlebanner{{\tt \textcolor{Red}{{\small Under Review -- distribute within CMU only.}}}}        % These are ignored unless
-%\preprintfooter{Distribute within CMU only.}   % 'preprint' option specified.
-
-\title{Active Type-Checking and Translation}
-
+\title{Active Typechecking and Translation}
+%
 %\authorinfo{~}{~}{~}
@@
         {\{comar, jonathan.aldrich\}@cs.cmu.edu}   
-
+%
 \maketitle
 \begin{abstract}
-% Active Code Completion
-% Active Typechecking and Translation
-% Active Type-Directed Source Code Prediction
-% Actively-Typed Parsing
-
-Researchers and domain experts typically describe new language-based abstractions as extensions of some existing language, but most statically-typed languages are actually \emph{monolithic}: they do not give users the ability to specify the semantics of new core types and operators from within.
-This choice has led to a proliferation of mutually incompatible standalone languages, each built around a small collection of privileged constructs.
-An alternative to this \emph{language-oriented} approach is to work in an extensible programming language where the compile-time behaviors determining the functionality of new core constructs are specified and implemented modularly within user libraries.
-Designing such a \emph{library-oriented} extensibility mechanism for statically-typed languages that is both safe and expressive is non-trivial.
+Researchers and domain experts typically describe new language-based abstractions as extensions of an existing  language, but most statically-typed languages are \emph{monolithic}: they do not give users the ability to specify the static and dynamic semantics of new types and operators from within.
+This has lead to a proliferation of mutually incompatible standalone languages, each built around a small collection of privileged constructs.
+An alternative to this \emph{language-external} approach is to work in an extensible programming language where the compile-time behaviors determining the functionality of new core constructs are provided directly within user libraries.
+Designing such a \emph{language-internal} extensibility mechanism that is both safe  and expressive is non-trivial.
 % -- extensions must not weaken the global safety properties of the language, compromise the integrity of the compilation process, or interfere with one other, while remaining flexible enough to admit a range of useful constructs.
 This paper introduces a mechanism called active type-checking and translation (AT\&T) that aims to address these  challenges.
-By relying upon type-level computation in a novel way, AT\&T admits user specification of a wide range of compile-time behaviors over a fixed grammar in a safe and modular manner.
-We discuss two points in the design space: (1) a simple calculus designed to distill the essential concepts and admit formal safety theorems, and (2) a fully-implemented language called Ace that we use to demonstrate the expressive power of AT\&T in several real-world domains, including scientific computing, security, functional programming and object-oriented programming.
+By building upon type-level computation in a novel way, AT\&T admits user specification of a wide range of compile-time behaviors over a fixed grammar in a safe and modular manner.
+We discuss two points in the design space: (1) a simple calculus designed to distill the essential concepts and admit formal safety theorems, and (2) a fully-implemented language called Ace that we use to demonstrate the expressive power of AT\&T across several domains, including scientific computing, security, functional programming and object-oriented programming.
 \end{abstract}
@@
 Programming languages have historically been specified and implemented monolithically.
-To introduce new primitive constructs, researchers or domain experts have developed a new language or a dialect of an existing language, with the help of tools like domain-specific language frameworks and compiler generators \cite{fowler2010domain}.
-Unfortunately, taking a so-called {\it language-oriented approach} \cite{journals/stp/Ward94}, where different languages are used for different components of an application, can lead to problems at language boundaries: a library's external interface must only rely on  constructs that can be expressed in all possible calling languages.
-This means that specialized invariants cannot be checked statically, decreasing reliability and performance.
-It also often requires that developers generate verbose and unnatural ``glue'' code, defeating a primary purpose of specialized languages: hiding these low-level details from end-user developers.
-
-Extensible programming languages promise to decrease the need for new standalone languages by providing more granular, language-based support for introducing new primitive constructs (that is, constructs that cannot be adequately expressed in terms of existing syntactic forms and primitive operations.) Developers would gain the freedom to choose those constructs that are most appropriate for their application domain and development discipline.
-Researchers would gain the ability to distribute new constructs for evaluation by a broader development community without requiring the approval of maintainers of mainstream languages, who are naturally risk-averse and uninterested in niche domains.
-
-A significant challenge that faces language extensibility mechanisms is in maintaining the overall safety properties of the language and compilation process in the presence of arbitrary combinations of user extensions.
-The mechanism must ensure that basic metatheoretic and global safety guarantees of the language cannot be weakened, that extensions are safely composable, and that type checking and compilation remains decidable.
-The correctness of an extension itself should be modularly verifiable, so that its users can rely on it for verifying and compiling their own code.
-These are the issues that we seek to address in this work.
-
-The approach we describe, {\it active type-checking and compilation} (AT\&T), makes use of type-level computation in a novel way.
-To review, in languages supporting type-level computation, the syntactic class of types is not simply declarative.
-Instead, it forms a programming language itself (the {\it type-level language}).
-Types themselves are one kind of value in this language, but there can be many others.
-To ensure the safety of type-level computations, {\it kinds} classify type-level terms, just as types classify expression-level terms.
-The simplest example of a language featuring type-level computation is Girard's System $\text{F}_{\omega}$ \cite{tapl}.
-In $\text{F}_{\omega}$, types have kind $\star$ and type-level functions have kinds like $\star\rightarrow\star$.
-A growing number of implemented languages now feature more sophisticated type-level languages (see Section 5).
-We emphasize that type-level computation occurs during compilation, rather than at run-time, because type-level terms that are used where types would normally be expected must be reduced to normal form before type-checking can proceed.
-
-In this work, we wish to allow extensions to strengthen the static semantics of our language.
-Naturally, extension specifications will also need to be evaluated during compilation and manipulate representations of types.
-This observation suggests that the type-level language may be able to serve directly as a specification language.
-In this paper, we show that this is indeed the case.
-By introducing some new constructs at the type-level, developers can specify the semantics of operators associated with newly-introduced families of primitive types with type-level functions.
-The compiler front-end invokes these functions to synthesize types for and assign meanings to expressions, by translation into a {\it typed internal language}.
-Unlike conventional metaprogramming systems, these {\it type-level specifications} do not directly manipulate or rewrite expressions.
-Instead, they examine and manipulate the types of these expressions.
-By using a sufficiently constraining kind system and incorporating techniques from typed compilation into the type-level language directly, the global safety properties of the language and compilation process can be guaranteed.
-In other words, users can only {\it increase} the safety of the language.
-
-We focus on extending the static semantics of a language with a fixed, though flexible, grammar.
-Techniques for extensible parsing have been proposed in the past (e.g.
-\cite{journals/entcs/BrabrandSV03}), and we conjecture that these can be made compatible with the mechanism described in this paper with some simple modifications, but we do not discuss this further here.
-We also focus on {\it functional}, rather than declarative, specifications of language constructs.
-Extracting a compiler from a declarative language specification (e.g.
-in Twelf \cite{twelf}) has not yet been shown practical, but we note that a future mechanism of this sort could safely target a language implementing the mechanism we discuss here.
-
-The organization and key contributions of this paper are:
-\begin{itemize}
-\item In Section 2, we develop a core calculus, $\lamA$, and give simple examples of language features that can be expressed with it.
+To introduce new primitive constructs, researchers or domain experts have developed a new language or a dialect of an existing language, often with the help of tools like domain-specific language frameworks and compiler generators \cite{fowler2010domain}.
+Unfortunately, using different languages for different components of an application, called a {\it language-oriented approach} \cite{journals/stp/Ward94}, has problems when crossing language boundaries: the external interface of a library must only use constructs that can be expressed in all possible calling languages.
+\TODO{(talk about typical design using a common language like the JVM)} Thus, any specialized invariants cannot be checked statically, decreasing reliability.
+It also often requires that developers generate unnatural ``glue'' code, impacting performance and defeating a primary purpose of specialized languages: abstracting these low-level details away from developers.
+
+Extensible programming languages promise to decrease the need for new standalone languages by providing more granular, language-internal support for introducing new primitive constructs (i.e.
+constructs that cannot be correctly and concisely expressed in terms of existing constructs).
+By using an extensible language, developers would gain the freedom to choose those constructs most suitable for their problem domain and development discipline.
+Researchers and domain experts would gain the ability to develop new constructs modularly and distribute them for evaluation by a broader development community without requiring either wholesale adoption of a new language or approval from the maintainers of widely-used languages, who are naturally risk-averse and are unlikely to cater to niche domains.
+The key difference between the language-external approach, characteristic of language frameworks,  and this language-internal approach, characteristic of extensible languages, is illustrated in Figure \ref{approaches}.
+\begin{figure*}
+\begin{center}
+\includegraphics[scale=0.5]{approaches.pdf}
+\end{center}
+\vspace{-20px}
+\caption{\small (a) With the language-oriented approach, novel constructs are packaged into separate languages.
+Users can only safely and naturally call into languages consisting of common constructs.
+(b) With the library-oriented approach, there is one language and novel constructs are packaged as normal libraries.
+Thus, interoperability is not a problem.
+in monolithic systems, this approach is less expressive.\label{approaches}}
+\end{figure*}
+
+A number of criteria constrain potential language-internal extensibility mechanisms, however.
+One key issue is that a candidate mechanism must maintain the safety properties of the language and compilation process in the presence of arbitrary combinations of user extensions.
+The mechanism must ensure that basic metatheoretic and global safety guarantees of the language cannot be weakened, that extensions are safely composable, and that typechecking and compilation remain decidable.
+The correctness of extensions themselves should be modularly verifiable, so that developers can rely on them for verifying and compiling their own code.
+At the same time, the extensibility mechanism should be capable of expressing a variety of language constructs naturally (from the perspective of the developers using the constructs).
+For extension developers, using the extension mechanism should be comparable in difficulty to using DSL frameworks and other language-external tools.
+These criteria guide the work that we describe here.
+
+We will begin by considering the typical compiler front-end, consisting of a typechecker and an initial translator to a high-level intermediate representation (e.g.
+bytecode for a virtual machine).
+The extensibility problem considered in this context can be cast as an instance of the \emph{expression problem} first examined by Reynolds \cite{reynoldsexpressionproblem} and so-named by Wadler \cite{expressionproblem}.
+Due to the particular structure of this domain, it is possible to sidestep the problem by delegating to functions associated with the internal representation of types, a form of \emph{inversion of control}.
+
+Building an extensible compiler front-end by exposing these representations via some API is a tempting approach, but it remains a  \emph{language-external} one.
+That is, a compiler that allows for new language constructs (by any mechanism) creates, \emph{de facto}, a new language (an underappreciated fact in many language communities!)  Thus, programs compiled against different collections of extensions may be mutually incompatible with one another.
+Moreover, such extensions cannot be used by different compilers for the same language.
+
+All is not lost, however.
+It can be observed that types are represented both within the compiler front-end and within the language itself.
+In most languages, the mechanisms for defining and referring to types are purely declarative.
+In languages featuring support for \emph{type-level computation}, however, types are a form of value that can be manipulated in various ways at compile-time via an internal \emph{type-level language} \cite{tlc}.
+This observation forms the foundation of the extensibility mechanism we describe here, called \emph{active typechecking and translation} (AT\&T), so named because it can be seen as a refinement of the general concept of \emph{active libraries}  proposed by veldHuizen \cite{activelibraries} (see Section \ref{relatedwork} for further discussion).
+
+We will introduce AT\&T by way of two artifacts: a fully-implemented language, Ace, and a minimal type-theoretic formulation, $\lamA$.
+Ace demonstrates the practicality of AT\&T across a number of problem domains, beginning with low-level, performance-critical programming (targeting the OpenCL language and compilation infrastructure) and then demonstrate support for language-internal extensions that add a variety of higher-level  abstractions in Section \ref{extensions}.
+Ace, as an implemented prototype that makes some concessions for the sake of usability, does not readily admit formal safety theorems.
+For this reason, we introduce $\lamA$, a terse and somewhat more limited formalism based on a typed lambda calculus with type-level computation of higher kind.
+This calculus captures the core elements of the   mechanism and admits several important safety theorems.
+
+%The Ace grammar is a nearly-complete subset of Python's  and we show that it is flexible enough to support all of the primitives of the C99 and OpenCL languages in a natural manner, when aided by a form of type inference that we describe.
+We further supplement these primitives with higher-level user-defined primitives, examples of which we discuss in Section \ref{abstractions}.
+Notably, all of these primitives are imported via the normal library loading mechanism; they are not defined within domain-specific languages, nor as extensions for a particular compiler implementation.
+%
+
+%The organization and key contributions of this paper are:
+%\begin{itemize}
+%\item In Section 2, we develop a core calculus, $\lamA$, and give simple examples of language features that can be expressed with it.
 We formalize the compiler front-end and state several lemmas that lead to useful safety theorems for the compiler and language as a whole.
 We show how AT\&T requires that the language provide a solution to a type-level variant of Wadler's expression problem \ref{Wadler}.
-\item In Section 3, we briefly introduce the Ace programming language, which is based fundamentally on an  elaboration of AT\&T that supports a richer set of syntactic forms and a variant of type inference.
+%\item In Section 3, we briefly introduce the Ace programming language, which is based fundamentally on an  elaboration of AT\&T that supports a richer set of syntactic forms and a variant of type inference.
 It uses object-oriented inheritance to solve the type-level expression problem.
 A number of practical extensions have been written using Ace, including a complete implementation of the OpenCL type system (based on C99) as a library.
-\item In Section 4, we briefly describe another point in the design space, a language design we call Birdie.
+%\item In Section 4, we briefly describe another point in the design space, a language design we call Birdie.
 Birdie lifts an extension of the Gallina language, used by the Coq proof assistant, into the type level (leading to a language with dependent kinds).
@@
 The expression problem is solved using a constrained formulation of open data types, rather than using object-oriented inheritance.
-\item In Section 5 we compare AT\&T to previous work on extensible languages and compilers, metaprogramming systems and formal specification languages and conclude in Section 6 with a discussion of future work.
+%\item In Section 5 we compare AT\&T to previous work on extensible languages and compilers, metaprogramming systems and formal specification languages and conclude in Section 6 with a discussion of future work.
+%\end{itemize}
+
+\section{Inversion of Control in a Compiler}\label{inversion}
+
+\section{Structure and Usage}
+%Ace is an {\em actively-typed, language-integrated compilation environment}. Let us begin by defining each of these terms. 
+%A {\em compilation environment} is a system that allows developers to programmatically control aspects of compilation directly.
+The language used to do so is called the {\em metalanguage} while the grammar that programs themselves are written in is called the {\em source grammar}.
+The metalanguage of Ace is Python and the source grammar is a significant subset of Python's own grammar.
+Programs and metalanguage code can be interleaved within the same file, so Ace is {\em language-integrated}.
+Ace programs are statically-typed, and types themselves are objects in the metalanguage.
+They can be equipped by developers with methods that control how operators associated with them are typechecked and compiled.
+Because of the relationship between this mechanism and the idea of {\em active libraries} \cite{activelibraries}, discussed further in Section \ref{related}, we call them {\em active types}, and thus Ace is {\em actively-typed}.
+\begin{figure}
+\lstinputlisting{hello.py}
+\caption{\texttt{[hello.py]} A basic Ace program demonstrating the two-phase structure of Ace programs and libraries.}
+\label{hello}
+\end{figure}
+\begin{figure}
+\begin{lstlisting}[style=Bash]
+$ `acec hello.py`
+Hello, compile-time world!
+Goodbye, compile-time world!
+$ `cat hello.cl`
+__kernel void main() {
+    printf("Hello, run-time world.");
+}\end{lstlisting}
+\caption{Compiling \texttt{hello.py} using the \texttt{acec} compiler.}
+\label{helloout}
+\end{figure}
+
+\subsection{Example 1: Hello, World!}
+To demonstrate the somewhat unconventional structure of Ace programs and libraries, we begin in Listing \ref{hello} with a variant of the canonical ``Hello, World!'' example written using the \verb|OpenCL| module.
+We note that this module, which we use throughout the paper in examples, is simply a library like any other.
+The core of Ace gives no special treatment to it; it is distributed with Ace for convenience.
+Aspects of its implementation will be described in Section \ref{att}.
+
+Perhaps the most immediately apparent departure from the standard ``Hello, World!'' example comes on lines 3 and 10 of Listing \ref{hello}, which contain \verb|print| statements that are executed at {\em compile-time}.
+Indeed, Ace programs are Python scripts at the top-level, rather than a list of declarations as in most conventional languages.
+In other words, Python is the {\em compile-time metalanguage} of Ace; Ace programs are embedded within Python.
+A consequence of this choice is that Ace can leverage Python's well-developed package system and  distribution infrastructure directly (Line 1).
+
+The \verb|main| function defined on Lines 5-7 contains a single call to the \verb|printf| primitive, imported from the \verb|ace.OpenCL| module on Line 1.
+The function is introduced using the \verb|@OpenCL.fn| decorator, which indicates that it is a statically-typed Ace function containing run-time logic targeting the \verb|OpenCL| backend, also imported on Line 1.
+Without this decorator, the function would simply be a conventional Python function that could be called only at compile-time (doing so would fail here: \verb|printf| is not a Python primitive.)
+
+On Line 8, this {\em generic function} defined on Lines 5-7, is compiled to produce a {\em concrete function} that we also identify as \verb|main| (overwriting the previous definition).
+The distinction between generic and concrete functions will be explained in our next example.
+For now, we note that only concrete functions are exported when running the Ace compiler, called \verb|acec|, shown operating at the shell in Listing \ref{helloout}.
+The \verb|acec| compiler operates as follows:
+\begin{enumerate}
+\item Executes the provided Python module (\verb|hello.py|)
+\item Produces source code associated with any concrete functions (functions produced using the \verb|compile| method on Line 8 of Listing \ref{hello}) that are in the top-level environment.
+Each backend may produce one or more files (here, \verb|hello.cl| is produced).
+\end{enumerate}
+
+Using the \verb|acec| executable is simply a convenience; the generated code is also available as an attribute of the concrete function immediately after it has been defined, accessed as \verb|main.code|.
+We will show in Section \ref{pyopencl} that, for backend languages with Python bindings, such as OpenCL and C, generic functions can be executed directly, without the intermediate compilation step, if desired.
+
+\subsection{Example 2: Higher-Order Map for OpenCL}
+\begin{figure}
+\lstinputlisting{listing3.py}
+\caption{[\texttt{listing\ref{map}.py}] A generic data-parallel higher-order map function written using the OpenCL user module.}
+\label{map}
+\end{figure}
+
+\begin{figure}
+\lstinputlisting{listing4.py}
+\caption{[\texttt{listing\ref{mapadd5dbl}.py}] The generic \texttt{map} function compiled to map the \texttt{add5} function over two  types of input.}
+\label{mapadd5dbl}
+\end{figure}
+
+\begin{figure}
+\lstinputlisting[style=OpenCL]{listing5.cl}
+\caption{[\texttt{listing\ref{mapout}.cl}] The OpenCL code generated by running \texttt{acec listing2.py}.}
+\label{mapout}
+\end{figure}
+
+The ``Hello, World!'' example demonstrates the basic structure of Ace programs, but it does not require working with types.
+Listing \ref{map} shows an imperative, data-parallel map primitive written using the OpenCL library introduced above.
+In OpenCL, users can define functions, called {\em kernels}, that execute across thousands of threads.
+Each kernel has access to a unique index, called its {\em global id}, which can be used by the programmer to ensure that each thread operates on different parts of the input data (Line 5).
+The \verb|map| kernel defined in Listing \ref{map} applies a transfer function, \verb|f|, to the element of the input array, \verb|input|, corresponding to its global id.
+It writes the result of this call into the corresponding location in the provided output array, \verb|output|.
+
+As above, \verb|map| is a {\em generic function} (specifically, an instance of the class \verb|ace.GenericFn|).
+This means that it's arguments have not been assigned types.
+Indeed, the functionality given by the \verb|map| definition is applicable to many combinations of types for \verb|input| and \verb|output| and functions, \verb|f|.
+In this sense, \verb|map| is actually a family of functions defined for all types assignments for \verb|input|, \verb|output| and \verb|f| for which the operations in the function's body are well-defined.
+
+Running \verb|acec listing3.py| would produce no output, however.
+To create a {\em concrete function} (an instance of the class \verb|ace.ConcreteFn|) that can be emitted by the compiler, types must be assigned to each of the arguments of any externally callable functions.
+Listing \ref{mapadd5dbl} shows how to use the \verb|compile| method to specialize \verb|map| in two different ways to apply the \verb|add5| function, defined on Lines 4-6, to arrays  that reside in global memory (OpenCL has a notion of four different memory spaces).
+Line 9 produces a version specialized for arrays of \verb|double|s and Line 10 produces a version for arrays of \verb|int|s.
+The output of compilation is shown in Figure \ref{mapout}.
+
+\subsection{Types as Metalanguage Objects}
+The types of the arguments of the functions being compiled on Lines 4.9 and 4.10, if written directly in OpenCL as in Listing \ref{mapout}, are \verb|__global double*| and \verb|__global int*|, respectively.
+The corresponding types in Ace are \verb|global_ptr(double)| and \verb|global_ptr(int)|, abbreviated for convenience as \verb|A| and \verb|B| respectively on Line 8 (note that here, standard variable assignment in the metalanguage is serving the role that a \verb|typedef| declaration would fill in a C-like language.)
+
+\textbf{Types in Ace are values in the metalanguage}, Python.
+More specifically, types are {\em object instances} of user-defined classes that inherit from the Ace-provided \verb|ace.Type| class.
+For example, \verb|global_ptr(double)| is an instance of \verb|ace.OpenCL.GlobalPtrType| instantiated with the target type, \verb|double|, as a constructor argument.
+The types \verb|double| and \verb|int| (imported from \verb|ace.OpenCL| on Line 4.2) are instances of \verb|ace.OpenCL.FloatType| and \verb|ace.OpenCL.IntegerType|, respectively.
+As we will describe further in Section \ref{att}, this notion of types as metalanguage objects is a key to Ace's flexibility.
+
+\subsection{Type Propagation and Higher-Order Functions}
+The type assigned to the third argument, \verb|f|, on Lines 4.9 and 4.10, is given as \verb|add5.ace_type|.
+The \verb|ace_type| attribute of a generic function is an instance of \verb|ace.GenericFnType|, the type of Ace generic functions, that corresponds to that particular function, \verb|add5| in this case.
+Generic functions are compiled automatically when they are called from another function.
+That is, when the compiler encounters the call to \verb|f| inside \verb|map| when compiling \verb|map_add5_double|, it compiles a version of \verb|add5| specialized to the \verb|double| type (seen on Line 5.3), and similarly with the \verb|int| type when compiling \verb|map_add5_int| (on Line 5.16, automatically given a unique name to avoid conflicts).
+This mechanism is called {\em type propagation}.
+In other words, we did not need to use \verb|add5.compile(double)| before compiling \verb|map_add5_dbl| (although we could if we would like).
+Only functions that are never called in the process of compiling other functions in a module need type information explicitly provided, whereas \verb|add5| is a function that is only called within \verb|map| in Listing \ref{mapadd5dbl}.
+
+We note that this scheme allows for a form of higher-order functional programming even when targeting languages, like OpenCL, that have no support for higher-order functions (OpenCL, unlike C99, does not even support function pointers).
+This works because the \verb|ace.GenericFnType| for one function, such as \verb|add5|, is not the same as the \verb|ace.GenericFnType| for a superficially similar function, such as \verb|add6| (defined as one would expect).
+To put it in type theoretic terms, \verb|ace.GenericFnType|s are singleton types, {\em uniquely inhabited} by a single generic function, and thus it is impractical to use them as full first-class values (i.e.
+they cannot put into a run-time array.) In practice, this is rarely a problem, particularly in parallel programming where explicit specialization is already common in order to avoid the potential performance penalty associated with a function pointer dereference per call.
+We note that fully-featured first-class functions can be implemented when targeting a backend that supports them, such as C99, or simulated by using a jump table implementation in OpenCL, but we do not discuss this further due to lack of space.
+
+\subsection{Type Inference}
+On Line 5 in the generic \verb|map| function in Listing \ref{map}, the variable \verb|gid| is initialized with the result of calling the \verb|get_global_id| primitive (the argument, \verb|0|, is not important for our purposes.)  Note that the type for the \verb|gid| variable is never listed explicitly.
+This is because Ace supports a form of {\em whole-function type inference}.
+In this case, \verb|gid| will be assigned type \verb|size_t| because that is the return type of \verb|get_global_id| (as defined in the OpenCL specification \cite{opencl}, which the \verb|ace.OpenCL| module follows.) The result can be observed on Lines 11 and 24 in Listing \ref{mapout}.
+
+Inference is not restricted within single assignments, as in the \verb|map| example.
+Multiple assignments to the same variable with values of differing types, or multiple return statements, can be unified such that the variable or return type is given a common supertype.
+For example, in the \verb|threshold_scale| function defined in Listing \ref{inference}, the variable \verb|y| in the first branch of the conditional is assigned the \verb|int| literal \verb|0|.
+However, in the second branch of the loop, its type depends on the types of \verb|x| and \verb|scale|.
+We show two choices for these types on Lines 10 and 11.
+However, type inference correctly unifies these two types according to OpenCL's C99-derived rules governing numeric types (which are defined by the user in the \verb|OpenCL| module, as we will describe in Section \ref{att}).
+We test this programmatically on Lines 12 and 13.
+Note that this example would also work correctly if the assignments to \verb|y| were replaced with \verb|return| statements (in other words, the return value of a function is treated as an assignable for the purpose of type inference).
+
+\begin{figure}
+\lstinputlisting{listing6.py}
+\caption{\texttt{[listing6.py]} A function demonstrating whole-function type inference when multiple values with differing types are assigned to a single variable.}
+\label{inference}
+\end{figure}
+
+\subsection{Annotation and Extension Inference}
+In addition to type annotations, OpenCL normally asks for additional annotations in a number of other situations.
+Users can annotate functions that meet certain requirements to be callable from the host with the \verb|__kernel| attribute.
+The \verb|Ace.OpenCL.OpenCL| backend is able to check these requirements and add this annotation automatically.
+Several types (notably, \verb|double|) and specialized functions require that an OpenCL extension be enabled with a \verb|#pragma| when used.
+The OpenCL backend automatically detects many of these cases as well and adds the appropriate \verb|#pragma| declaration.
+An example of this feature can be seen on Line 5.1, where the use of the \verb|double| type triggers the insertion of an appropriate \verb|#pragma| automatically.
+Ace is designed to allow backends to observe the results of the type checking process to support this form of inference.
+
+\subsection{Metaprogramming in Ace}
+\begin{figure}
+\lstinputlisting[commentstyle=\color{mauve}]{listing7.py}
+\caption{[\texttt{listing7.py}] Metaprogramming with Ace, showing how to construct generic functions from both strings and abstract syntax trees, and how to manipulate syntax trees at compile-time.}
+\label{metaprogramming}
+\end{figure}
+Metaprogramming refers to the practice of writing programs that manipulate other programs.
+There are a number of use cases for this technique, including domain-specific optimizations and code generation for programs with a repetitive structure that cannot easily be captured using available abstractions \cite{pyopencl}.
+OpenCL in particular relies on code generation as a fundamental mechanism, which is cited as justification for its lack of support for higher-order programming.
+Ace supports programmatic compilation and higher-order constructs, as described above, and a flexible language extension mechanism, which we describe below, so several use cases for metaprogramming have been eliminated.
+However, cases where this form of metaprogramming include programmatic function specialization (Listing \ref{metaprogramming}) and modular simulation orchestration, described in Section \ref{clegans}.
+
+On Lines 4-7 of Listing \ref{metaprogramming}, an Ace function is constructed from a string containing its source using the \verb|from_source| variant of the \verb|OpenCL.fn| method.
+It can also be constructed directly from an abstract syntax tree (AST), as implemented by the Python standard \verb|ast| package, using the \verb|from_ast| variant of \verb|fn|, demonstrated on Line 10.
+The AST here is generated programmatically by calling the \verb|specialize| function, which produces a copy of the syntax tree of the \verb|plus| function with the argument, \verb|b|, eliminated and its uses replaced with a constant, \verb|5|.
+This transformation as well as some others are distributed in \verb|ace.astx| for convenience.
+\subsection{Direct Invocation from Python}\label{direct}
+
+\begin{figure}
+\lstinputlisting{listing8.py}
+\caption{[\texttt{listing\ref{py}.py}] A full OpenCL program using the \texttt{Ace.OpenCL} Python bindings, including data transfer to and from a device and direct invocation of a generic function, \texttt{map}, as a kernel without explicit compilation.}
+\label{py}
+\end{figure}
+
+As discussed in the Introduction, a common workflow for professional end-users involves the use of a high-level scripting language for overall workflow orchestration and small-scale data analysis and visualization, paired with a low-level language for performance-critical sections.
+Python is already widely used by professional end-users as a high-level scripting language and also features mature support for calling into code written in low-level languages.
+Developers can call into native libraries using its foreign function interface (FFI),  or by using a wrapper library like \verb|pycuda| for code compiled with CUDA, a proprietary language similar to OpenCL  specifically targeting nVidia GPU hardware.
+Although CUDA's compilers are separate executables on the system, the OpenCL language was designed for this workflow, in that it exposes the compiler directly as an API.
+The \verb|pyopencl| module exposes this API as well as the OpenCL memory management API to Python.
+With both \verb|pyopencl| and \verb|pycuda|, developers generate source code as strings, compile it programmatically, then execute it using the run-time APIs that each library provides \cite{pyopencl}.
+
+Ace supports this workflow as an alternative to using the \verb|acec| compiler as described above to generate  source code directly from the shell.
+For the OpenCL backend, these bindings are exposed as a wrapper on top of \verb|pyopencl| called \verb|Ace.OpenCL.bindings|.
+Both generic functions and concrete functions written for a backend that supports the direct execution interface (thus far, \verb|OpenCL| and \verb|C99|) can be called like regular Python functions.
+An example of this for the generic \verb|map| function defined in Listing \ref{map} is shown in Listing \ref{py}, with the call itself on Lines 9-10.
+The first two arguments to \verb|map| are OpenCL buffers, generated using a simplified wrapper to the \verb|pyopencl| APIs on Lines 7-8.
+This wrapper associates type information with each buffer, similarly to \verb|numpy|, and this is used to implicitly compile \verb|map| as appropriate the first time it is called for any given combination of input types.
+Explicit calls to the \verb|compile| method, as we have been showing thusfar, are unnecessary if using this method of invocation.
+The final two keyword arguments on Line 10 are parameters that OpenCL requires for execution that determine the number of threads (called the {\em global size}) and thread grouping (the {\em local size}).
+
+By way of comparison, the same program written using the OpenCL C API directly is an order of magnitude larger and correspondingly more complex.
+A full implementation of the logic of \verb|map| written using the \verb|pyopencl| bindings and metaprogramming techniques as described in \cite{pyopencl} is twice as large and significantly more difficult to comprehend than the code we have shown thus far.
+Not shown are several additional conveniences, such as delegated kernel sizing and \verb|In| and \verb|Out| constructs that can reduce the size and improve the clarity of this code further; due to a lack of space, the reader is referred to the language documentation for additional details on these features.
+
+\section{Ace for Researchers}
+Thus far, we have been largely discussing the OpenCL module in our examples.
+However, Ace gives no preferential treatment to this module; it is implemented entirely using the user-facing mechanisms described in this section.
+A C99 module has also been developed (which, due to its similarity to OpenCL, shares many of its implementation details), but we do not discuss it further in this paper.
+Extensions complementing these are described in Section C below.
+
+Most programming languages are {\em monolithic} -- the set of available primitives is determined by the language designers, and users must combine these primitives to produce any desired run-time behavior.
+Although many very general primitives have been developed (e.g.
+object systems and algebraic datatypes), these can be insufficient in specialized domains where developers and researchers need fine control over how certain operations are type checked and translated.
+High-performance computing is an example of such a field, since both correctness and performance have been difficult to achieve in general, and are topics of active research.
+As discussed in the Introduction, the proliferation of parallel programming languages, rather than library-based abstractions, indicates that researchers often find the abstractions available in existing languages insufficient for their needs.
+
+To address these use cases, Ace has been designed to be fundamentally {\em extensible}, rather than monolithic.
+Users can introduce new primitive types and operations and fully control how they are typechecked and translated.
+The backend target of translation can also be specified modularly by the user.
+Because Ace libraries can contain compile-time logic, written in the metalanguage as described in the previous section, these primitive definitions can be distributed as modules, rather than as extensions to particular compilers or domain-specific languages.
+
+\subsection{Active Typechecking and Translation (\ATT)}
+We now explain how new primitive types and operators can be defined in Ace.
+When the compiler encounters an expression, such as \verb|input[gid]|, it must first verify its validity by assigning it a type, then translate the expression to produce an expression in a target language.
+Rather than containing fixed logic for this, however, the Ace compiler defers this responsibility to the {\it type} of a subexpression, such as \verb|input|, whenever possible, according to a fixed {\em dispatch protocol} for each syntactic form.
+Below are examples of the rules that comprise the Ace dispatch protocol.
+Due to space constraints, we do not list the entire dispatch protocol, which contains a rule for each possible syntactic form in the language.
+\begin{itemize}
+\item Responsibility over a {\bf unary operation} like \verb|-x| is handed to the type assigned to the operand, \verb|x|.
+\item Responsibility over {\bf binary operations} is first handed to the type assigned to the left operand.
+If it indicates that it does not understand the operation, the type assigned to the right operand is handed responsibility, with a different method call\footnote{Note that this operates similarly to the Python run-time operator overloading protocol; see Related Work.}.
+\item Responsibility over {\bf attribute access} (\texttt{obj.attr}) and {\bf subscript access}, (\texttt{obj[idx]}) is handed to the type assigned to \texttt{obj}.
+\item {\color{red} Talk about multiple assignment here I think}
 \end{itemize}
-
-\section{Type-Level Specifications in $\lamA$}
-\subsection{Example: Natural Numbers in $\lamA$} % G\"{o}del's $T$ 
-We begin with a simple calculus with no primitive notion of natural numbers, nor any more general notion of an inductive data type.
-We can, however, concretely specify both the static and dynamic semantics of natural numbers, including the natural recursor of G\"{o}del's $T$ \cite{tapl}, using type-level specifications.
-Let us begin in Figure 1 with the type $\tvar{nat}$ and its constructors, $\tvar{z}$ and $\tvar{s}$.
 \begin{figure}
-$\tfam{NAT}{\kunit}{
-	\tlam{\tof{\tvar{self}}{\kType{\fvar{NAT}}}}{\titype{\dint}}}{(\\
-	~~~~
-	\topp{\tops{z}{\tlam{\tof{\tvar{self}}{\kTypeBlur}}{\tconst{\tden{\titerm{0}}{\tvar{self}}}}}}{
-	\\~~~~
-	\tops{s}{\tlam{\tof{\tvar{self}}{\kTypeBlur}}{\tOp{
-		\tlam{\tof{\tvar{d1}}{\kDen}}{
-		\\~~~~~~~~~\tifeq{\ttypeof{\tvar{d1}}}{\tvar{self}}{\tden{\tvalof{\tvar{d1}}}{\tvar{self}}}{\terr}}}}}}
-	)}{\Theta}{{\textsf{let~}}{\text{nat}=\ttype{\tunit}{\fvar{nat}}}{\textsf{~in~}}{x}}$
-\caption{Specifying natural numbers in $\lamA$. This term is referred to as $\psit{nat}$.}
+\lstinputlisting{listing9.py}
+\caption{\texttt{[listing9.py]} A portion of the implementation of OpenCL pointer types implementing subscripting logic using the Ace extension mechanism.}
+\label{pointers}
 \end{figure}
-\begin{figure*}
-\small
-$$\begin{array}{rccl}
-\textbf{expressions} 				&	e	&	::=	&	\evar{x} \pipe 
-														\elam{\evar{x}}{\psi}{e} \pipe 
-														\eapp{e_{1}}{e_{2}} \pipe
-												\eop{\psitype}{op}{
-															\splat{\psi}{1}{m}
-														}{
-
-														} \pipe
-														\eop{\psitype}{op}{
-															\splat{\psi}{1}{m}
-														}{
-  												    		\splat{e}{1}{n}
-														} \\
-									& 		&		& 	\\
-
-\textbf{type-level specifications} 	& \psi 	& ::= 	& 	\tvar{t} \pipe 
-														\tlam{\splatTwo{\tvar{t}}{{:}\kappa}{1}{n}}{\psi} \pipe 
-														\tapp{\psi}{\splat{\psi}{1}{n}} \pipe
-														\tifeq{\psi_{0}}{\psi_{1}}{\psi_{2}}{\psi_{3}} \\
-
-\text{standard terms}	 			& 		& \pipe	& 	\tunit \pipe 
-														\tpair{\psi_{1}}{\psi_{2}} \pipe 
-														\tfst{\psi} \pipe 
-														\tsnd{\psi} 
-														\\
-
-\text{type families}				&		& \pipe	& 	\tfam{fam}{\kappaidx}{\psirep}{\theta}{\Theta}{\psi}\pipe
-														\tfamcase{\psi}{\phi}{\psi_0}{\psi_1}\\
-
-\text{types} 						& 		& \pipe	& 	\ttypestd \pipe 
-														\tidx{\psitype} \pipe
-														\trepof{\psitype}
-														\\
-
-\text{operator definitions} 		& 		& \pipe & 	\tconst{\psi} \pipe 
-														\tOp{\psi} \\
-
-\text{denotations} 				& 		 & 	\pipe	&	\tden{\psiiterm}{\psitype} \pipe 
-														\tvalof{\psiden} \pipe
-														\ttypeof{\psiden} \pipe
-														\terr \\
-
-\text{internal language}		&		&	\pipe	&	\titerm{\mu} \pipe \titype{\delta}\\
-
-\text{programs}					&		&	\pipe	&	\tprog{e}\\
-
-\text{family specifications}	&	\phi		&	::= & \fvar{fam} \pipe \tfamSpecStd\\
-\text{operator lists}		&	\theta	&	::= &	\topsempty \pipe 
-												\topp{\theta}{\tops{op}{\psi}}\\
-
-\textbf{kinds} 					& \kappa	&	::=	&	\karrow{\splat{\kappa}{1}{n}}{\kappa} \pipe 
-												\kunit \pipe 
-												\kpair{\kappa_{1}}{\kappa_{2}} \pipe 
-												\kType{\eta} \pipe 
-												\kTypeBlur \pipe
-												\kOp{n} \pipe
-												\kDenk{\eta} \pipe 
-												\kDen \pipe 
-												\kIType \pipe \kITerm \pipe
-												\kProg 
-												\\
-
-\text{family signatures}				& \eta		& 	::=	& \kFamStd\\
-\text{operator list signatures}			& \Theta	&	::=	&	\cdot \pipe \Theta; \tvar{op}: \kappa\\
-											 							&		&		&	\\
-\textbf{internal terms} 				& 	\mu	&	::=	&	\ivar{u} \pipe 
-												\ilam{\ivar{u}}{\delta}{\mu} \pipe 
-												\ifix{\ivar{f}}{\delta}{\mu} \pipe 
-												\iapp{\mu_{1}}{\mu_{2}} \pipe 
-												\iIfEq{\mu_{0}}{\mu_{1}}{\mu_{2}}{\mu_{3}}\\
-							& 		& 	\pipe	&	\iintlit \pipe 
-												\iop{\mu_{1}}{\mu_{2}} \pipe 
-												\ipair{\mu_{1}}{\mu_{2}} \pipe 
-												\ifst{\mu} \pipe
-												\isnd{\mu} \pipe
-												\iup{\psi} 
-												\\
-\textbf{internal types}			&	\delta	&	::=	&    \darrow{\delta_1}{\delta_2} \pipe
-												\dint \pipe
-												\dpair{\delta_1}{\delta_2} \pipe
-												\dup{\psi}\\
-%\text{cases}					&\tcasev 	&	::=	&	\caseZ{\splat{\kappa}{1}{n}}{\psi} \pipe 
-%												\caseS{\tcasev}{\tcasev}\\
-%							&		&		&	\\
-\end{array}$$
-\caption{Syntax of $\lamA$. $\mathbb{Z}$ denotes integer literals and $\klabel{label}$ denotes any label.}
-\end{figure*}
-\begin{figure*}
-\begin{mathpar}
-\inferrule{
-	\overbrace{\tentailst{\cdot}{\psiprog}{\kProg}}^{\text{\small kind checking}}{ }\\
-	\overbrace{\tevalsX{\psiprog}{\tprog{e}}{\noprob}}^{\text{\small specification normalization}}{ }\\
-	\overbrace{\translates{\cdot}{e}{\mu}{\ttypestd}}^{\text{\small verification \& translation}}{ }
-}{
-%	\compiles{\psiprog}{\mu}{\psiidx}{\kappa}{\psirep}
-	\compiles{\psiprog}{\mu}{\ttypestd}
-}
-\end{mathpar}
-\caption{Central compilation judgement of $\lamA$.}
-\end{figure*}
-%\noindent
-%
-%The \textsf{def} construct is the type-level analog of the standard \textsf{let} construct.
-Indeed, the entire term consists of type-level constructs.
-The definition of $\tvar{nat}$ says that it is a type indexed by a uniquely labeled unit (that is, it is a simple type) that will be represented in the internal language by integers.
-The constructor $\tvar{z}$ is declared as a constant, denoted by the value $0$ and assigned the type $\tvar{nat}$.
-The successor operator $\tvar{s}$ first checks its argument to ensure that it is a $\tvar{nat}$ (which requires first checking the index's kind to ensure that it has the correct label using the $\sf{idxcase}_{D}$ statement).
-If so, it returns a denotation that assigns the type $\tvar{nat}$ to an internal term that increments the value of the argument.
-Otherwise, the special denotation $\terr$ is returned.
-We will return to explain these constructs in more detail shortly.
-%
-%This term, which we will call $\psit{nat}$, is an open term because it has not been used to write a program yet.
-To use this specification to write a program, we substitute a \textsf{program} term containing an expression referring to these definitions for $\tvar{x}$.
-For example, the following program corresponds to the  number $1$:
-$$\psi_{1} := \subst{\tprog{
-	\eop{\text{nat}}{s}{}{\eop{\text{nat}}{z}{}{}}
-}}{\psi_{\text{nat}}}$$
-%\noindent
-%This program will be compiled to the internal term $1 + 0$ and assigned the type $\tvar{nat}$.
-This notion is formalized by the {\it central compilation judgement} of $\lamA$, given in Figure 3.
-In this case, we will be able to derive that
-$$
-\compiless{\psi_{1}}{1 + 0}{\tvar{nat}}
-$$
-%Compilation will not succeed for invalid terms, e.g.
-%\begin{eqnarray*}
-%\psi_{-} & := & \subst{\tprog{\econst{\tvar{s}}}}{\tvar{x}}{\psi_{\text{nat}}}\\
-%\psi_{-}' & := & \subst{\tprog{\eop{\tvar{s}}{\elam{\evar{n}}{\tvar{nat}}{\eop{\tvar{s}}{\evar{n}}}}}}{\tvar{x}}{\psi_{\text{nat}}}
-%\end{eqnarray*}
-%Indeed, compilation with $\psi_{\text{nat}}$ as given above admits all well-typed expressions of the simply-typed lambda calculus with natural numbers and no others, compiling $\tvar{nat}$s to internal terms that evaluate to their integer representations.
-%
-%\subsection{$\lamA$}
-\begin{figure*}
-$\fbox{\inferrule{}{\tentailsXt{\psi}{\kappa}}}$~
-~~~~$\fCtx ::= \cdot \pipe \fCtx, \fSpecStd$
-~~~~$\tvarCtx ::= \cdot \pipe \tvarCtx, \tof{\tvar{t}}{\kappa}$
-\begin{mathpar}
-\small
-\inferrule{
-	\tof{\tvar{t}}{\kappa} \in \tvarCtx
-}{
-  \tentailsXt{\tvar{t}}{\kappa}
-}~(\text{var}_\psi)
-
-\inferrule{ }{
-	\tentailsXt{\tunit}{\kunit}
-}~(\text{unit}_\psi)
-
-\inferrule{
-	\tentailsXt{\psi_{1}}{\kappa_{1}}\\
-	\tentailsXt{\psi_{2}}{\kappa_{2}}
-}{
-	\tentailsXt{\tpair{\psi_{1}}{\psi_{2}}}{\kpair{\kappa_{1}}{\kappa_{2}}}
-}~(\times_\psi)
-
-\inferrule{
-	\tentailsXt{\psi}{\kpair{\kappa_{1}}{\kappa_{2}}}
-}{
-	\tentailsXt{\tfst{\psi}}{\kappa_{1}}
-}~(\text{fst}_\psi)
-
-\inferrule{
-	\tentailsXt{\psi}{\kpair{\kappa_{1}}{\kappa_{2}}}
-}{
-	\tentailsXt{\tsnd{\psi}}{\kappa_{2}}
-}~(\text{snd}_\psi)
-
-\inferrule{
-  \kentailsX{\kappa_1}\\
-  \cdots\\
-  \kentailsX{\kappa_n}\\
-  \tentailst{\tCtxXT{\splatTwo{\tvar{t}}{{:}\kappa}{1}{n}}}{\psi}{\kappa}
-}{
-  \tentailsXt{\tlam{\splatTwo{\tvar{t}}{{:}\kappa}{1}{n}}{\psi}}{\karrow{\splat{\kappa}{1}{n}}{\kappa}}
-}~\text{($\lambda_\psi$)}
-
-\inferrule{
-  \tentailsXt{\psi}{\karrow{\splat{\kappa}{1}{n}}{\kappa}}\\
-  \tentailsXt{\psi_{1}}{\kappa_{1}}\\
-  \cdots\\
-  \tentailsXt{\psi_{n}}{\kappa_{n}}
-}{
-  \tentailsXt{\tapp{\psi}{\splat{\psi}{1}{n}}}{\kappa}
-}~(\text{app}_\psi)
-
-\inferrule{
-	\tentailsXt{\psi_{0}}{\kappa_{1}}\\
-	\tentailsXt{\psi_{1}}{\kappa_{1}}\\
-	\tentailsXt{\psi_{2}}{\kappa_{2}}\\
-	\tentailsXt{\psi_{3}}{\kappa_{2}}
-}{
-	\tentailsXt{\tifeq{\psi_{0}}{\psi_{1}}{\psi_{2}}{\psi_{3}}}{\kappa_{2}}
-}~(\text{cond}_\psi)
-
-\inferrule{
-	\tentailst{\tCtxXF{\fSpecStd}}{\tfamSpecStd}{\kFamStd}\\
-	\tentailst{\tCtxXF{\tof{\fvar{fam}}{\kFamStd}}}{\psi}{\kappa}
-}{
-	\tentailsXt{\tfamStd}{\kappa}
-}~(\text{fambind})
-
-\inferrule{
-	\tentailsXt{\psitype}{\kTypeBlur}\\
-	\tentailsXt{\phi}{\eta}\\
-	\tentailsXt{\psi_0}{\karrow{\kType{\eta}}{\kappa}}\\
-	\tentailsXt{\psi_1}{\kappa}
-}{
-	\tentailsXt{\tfamcase{\psitype}{\phi}{\psi_0}{\psi_1}}{\kappa}
-}~(\text{famcase-T})
-
-\inferrule{
-	\tentailsXt{\psiden}{\kDen}\\
-	\tentailsXt{\phi}{\eta}\\
-	\tentailsXt{\psi_0}{\karrow{\kDenk{\eta}}{\kappa}}\\
-	\tentailsXt{\psi_1}{\kappa}
-}{
-	\tentailsXt{\tfamcase{\psiden}{\phi}{\psi_0}{\psi_1}}{\kappa}
-}~(\text{famcase-D})
-
-\inferrule{
-\tentailsXt{\phi}{\kFamStd}\\
-\tentailsXt{\psiidx}{\kappaidx}
-}{
-	\tentailsXt{\ttype{\psiidx}{\phi}}{\kType{\kFamStd}}
-} ~(\text{Type}_I)
-
-\inferrule{
-	\tentailsXt{\psi}{\kType{\eta}}
-}{
-	\tentailsXt{\psi}{\kTypeBlur}
-}~(\text{Type-$\subseteq$})
-
-\inferrule{
-	\tentailsXt{\psi}{\kType{\kFamStd}}
-}{
-	\tentailsXt{\tidx{\psi}}{\kappaidx}
-}~(\text{idxof})
-
-\inferrule{
-	\tentailsXt{\psi}{\kTypeBlur}
-}{
-	\tentailsXt{\trepof{\psi}}{\kIType}
-}~(\text{repof})
-
-\inferrule{
-	\tentailsXt{\psi}{\kDen}
-}{
-	\tentailsXt{\tconst{\psi}}{\kOp{0}}
-}~(\text{const})
-
-\inferrule{
-	\tentailsXt{\psi}{\karrow{\kDen, \stackrel{n}{\cdots}, \kDen}{\kDen}}
-}{
-	\tentailsXt{\tOp{\psi}}{\kOp{n}}
-}~(\text{$n$-op})
-
-\inferrule{
-	\tentailsXt{\psiiterm}{\kITerm}\\
-	\tentailsXt{\psitype}{\kType{\eta}}
-}{
-	\tentailsXt{\tden{\psiiterm}{\psitype}}{\kDenk{\eta}}
-}~(\text{Den}_I)
-
-\inferrule{
-	\tentailsXt{\psiiterm}{\kITerm}\\
-	\tentailsXt{\psitype}{\kTypeBlur}
-}{
-	\tentailsXt{\tden{\psiiterm}{\psitype}}{\kDen}
-}~(\text{Den}_I\text{-$\subseteq$})
-
-\inferrule{
-	\tentailsXt{\psi}{\kDenk{\eta}}
-}{
-	\tentailsXt{\psi}{\kDen}
-}~(\text{Den-$\subseteq$})
-
-\inferrule{ }{
-	\tentailsXt{\terr}{\kDen}
-}~(\text{err})
-
-\inferrule{
-	\tentailsXt{\psi}{\kDenk{\eta}}
-}{
-	\tentailsXt{\ttypeof{\psi}}{\kType{\eta}}
-}~(\text{typeof})
-
-\inferrule{
-	\tentailsXt{\psi}{\kDenk{\eta}}
-} {
-	\tentailsXt{\tvalof{\psi}}{\kITerm}
-}~(\text{valof})
-\\
-\inferrule{
-	\ientails{\iMkCtx{\fCtx}{\tvarCtx}{\cdot}}{\mu}
-}{
-	\tentailsXt{\titerm{\mu}}{\kITerm}	
-}~(\text{iterm})
-
-\inferrule{
-	\tentailsX{\delta}
-}{
-	\tentailsXt{\titype{\delta}}{\kIType}
-}~(\text{itype})
-
-\inferrule{
-	\ientails{\iMkCtx{\fCtx}{\tvarCtx}{\cdot}}{e}
-}{
-	\tentailsXt{\tprog{e}}{\kProg}
-}~(\text{program})
-\end{mathpar}
-$\fbox{\inferrule{}{\tentailsXt{\phi}{\eta}}}$~
-\begin{mathpar}
-\small
-\inferrule{
-	\tof{\fvar{fam}}{\eta} \in \Sigma
-}{
-	\tentailsXt{\fvar{fam}}{\eta}
-}~(\text{var}_\phi)
-
-\inferrule{
-	\kentailsX{\kFamStd}\\
-	\tentailsXt{\psirep}{\karrow{\kappaidx}{\kIType}}\\
-	\tentailsXt{\theta}{\Theta}
-}{
-	\tentailsXt{\tfamSpecStd}{\kFamStd}
-}~(\text{family}_I)
-\end{mathpar}
-$\fbox{\inferrule{}{\tentailsXt{\theta}{\Theta}}}$
-\begin{mathpar}
-\small
-\inferrule{ }{
-	\tentailsXt{\cdot}{\cdot}
-}~(\text{no-ops})
-
-\inferrule{
-	\tentailst{\Sigma, \tof{\fvar{fam}}{\eta}~\Delta}{\theta}{\Theta}\\
-	\tentailst{\Sigma, \tof{\fvar{fam}}{\eta}~\Delta}{\psi}{\karrow{\kType{\eta}}{\kappa}}
-}{
-	\tentailst{\Sigma, \tof{\fvar{fam}}{\eta}~\Delta}{\topp{\theta}{\tops{id}{\psi}}}{\Topp{\Theta}{\Tops{id}{\kappa}}}
-}~(\text{op-seq})
-\end{mathpar}
-\caption{Kind checking rules for $\lamA$. All contexts have standard substructural properties (not shown).}
-\end{figure*}
-\begin{figure*}
-$\fbox{\inferrule{}{\kentailsX{\kappa}}}$~
-$\fbox{\inferrule{}{\kentailsX{\kSimple{\kappa}}}}$
-%~~~~$\$
-\begin{mathpar}
-\small
-\inferrule{
-	\kentailsX{\kappa_1}\\
-	\cdots\\
-	\kentailsX{\kappa_n}\\
-	\kentailsX{\kappa}
-}{
-	\kentailsX{\karrow{\splat{\kappa}{1}{n}}{\kappa}}
-}~(\text{$\rightarrow$}_\kappa)
-
-\inferrule{
-	\kentailsX{\kappa_1}\\
-	\kentailsX{\kappa_2}
-}{
-	\kentailsX{\kpair{\kappa_1}{\kappa_2}}
-}~(\text{$\times$}_\kappa)
-
-\inferrule{
-	\kentailsX{\kSimple{\kappa}}
-}{
-	\kentailsX{\kappa}
-}~(\text{simple})
-
-\inferrule{
-	\kentailsX{\kSimple{\kappa_1}}\\
-	\kentailsX{\kSimple{\kappa_2}}
-}{
-	\kentailsX{\kSimple{\kpair{\kappa_1}{\kappa_2}}}
-}~(\text{$\times_\kappa$-simple})
-
-\inferrule{ }{
-	\kentailsX{\kSimple{\kunit}}
-}~(\text{Unit}_\kappa)
-
-\inferrule{ }{
-	\kentailsX{\kSimple{\kTypeBlur}}
-}~(\text{Type-$\subseteq$}_\kappa)
-
-\inferrule{
-	\kentailsX{\eta}
-}{
-	\kentailsX{\kSimple{\kType{\eta}}}
-}~(\text{Type}_\kappa)
-
-\inferrule{ }{
-	\kentailsX{\kSimple{\kDen}}
-}~(\text{Den-$\subseteq$}_\kappa)
-
-\inferrule{
-	\kentailsX{\eta}
-}{
-	\kentailsX{\kSimple{\kDenk{\eta}}}
-}~(\text{Den}_\kappa)
-
-\inferrule{ }{
-	\kentailsX{\kSimple{\kIType}}
-}~(\text{IType}_\kappa)
-
-\inferrule{ }{
-	\kentailsX{\kSimple{\kITerm}}
-}~(\text{ITerm}_\kappa)
-
-\inferrule{ }{
-	\kentailsX{\kSimple{\kProg}}
-}~(\text{Program}_\kappa)
-
-\inferrule{ }{
-	\kentailsX{\kSimple{\kOp{n}}}
-}~(\text{$n$-Op}_\kappa)
-\end{mathpar}
-$\fbox{\inferrule{}{\kentailsX{\eta}}}$~
-\begin{mathpar}
-\small
-\inferrule{
-	\kentailsX{\kSimple{\kappaidx}}\\
-	\kentailsX{\Theta}
-}{
-	\kentailsX{\kFamStd}
-}~(\text{family}_\eta)
-\end{mathpar}
-$\fbox{\inferrule{}{\kentailsX{\Theta}}}$
-\begin{mathpar}
-\small
-\inferrule{~}{
-	\kentailsX{\cdot}
-}~(\text{No-Ops})
-%
-%\inferrule{
-%	\kentailsX{\Theta}
-%}{
-%	\kentailsX{\Topp{\Theta}{\Tops{id}{\kOp{n}}}}
-%}~(\text{(0, $n$)-Op-Seq})
-
-\inferrule{
-	\kentailsX{\Theta}\\
-	\kentailsX{\karrow{\splat{\kappa}{1}{m}}}{\kOp{n}}
-}{
-	\kentailsX{\Topp{\Theta}{\Tops{id}{\karrow{\splat{\kappa}{1}{m}}{\kOp{n}}}}}
-}~(\text{($m$, $n$)-Op-Seq})
-\end{mathpar}
-$\fbox{\inferrule{}{\eentailsX{e}}}$
-~~~~$\etvarCtx ::= \cdot \pipe \etvarCtx, \evar{x}$
-\begin{mathpar}
-\small
-\inferrule{
-	\evar{x} \in \etvarCtx
-}{
-	\eentailsX{\evar{x}}
-}~(\text{var-form}_e)
-
-\inferrule{
-	\tentailsXt{\psi}{\kTypeBlur}\\
-	\eentails{\fCtx}{\tvarCtx}{\etvarCtx, \evar{x}}{e}
-}{
-	\eentailsX{\elam{\evar{x}}{\psi}{e}}
-}~(\text{$\lambda$-form}_e)
-
-\inferrule{
-	\eentailsX{e_1}\\
-	\eentailsX{e_2}
-}{
-	\eentailsX{\eapp{e_1}{e_2}}
-}~(\text{app-form}_e)
-
-\inferrule{
-	\tentailsXt{\psitype}{\kType{\kFamStd}}\\
-	\tof{\tvar{id}}{\karrow{\splat{\kappa}{1}{m}}{\kOp{0}}}\in\Theta\\
-	\tentailsXt{\psi_1}{\kappa_1}\\
-	\cdots\\
-	\tentailsXt{\psi_m}{\kappa_m}
-}{
-	\eentailsX{\eop{\psitype}{\tvar{id}}{\splat{\psi}{1}{m}}{ }}
-}~(\text{$(m, 0)$-op-app-form}_e)
-
-\inferrule{
-	\tentailsXt{\psitype}{\kType{\kFamStd}}~~~~~~~~
-	\tof{\tvar{id}}{
-		\karrow{\splat{\kappa}{1}{m}}{
-			\kOp{n}
-		}
-	} \in \Theta~~~~~~~~
-	\tentailsXt{\psi_1}{\kappa_1}~~~~~~~~
-	\cdots~~~~~~~~
-	\tentailsXt{\psi_m}{\kappa_m}\\
-	\eentailsX{e_1}\\
-	\cdots\\
-	\eentailsX{e_n}
-}{
-	\eentailsX{\eop{\psitype}{\tvar{id}}{\splat{\psi}{1}{m}}{\splat{e}{1}{n}}}
-}~(\text{$(m, n)$-op-app-form}_e)
-\end{mathpar}
-$\fbox{\inferrule{}{\ientailsX{\mu}}}$
-~~~~$\itvarCtx ::= \cdot \pipe \itvarCtx, \ivar{u}$
-\begin{mathpar}
-\small
-\inferrule{
-	\ivar{u} \in \itvarCtx
-}{
-	\ientailsX{\ivar{u}}
-}~(\text{var-form}_{\mu})
-
-\inferrule{
-	\tentailsX{\delta}\\
-	\ientails{\iMkCtx{\fCtx}{\tvarCtx}{\itvarCtx}, \ivar{u}}{\mu}
-}{
-	\ientailsX{\ilam{\ivar{u}}{\delta}{\mu}}
-}~(\text{$\lambda$-form}_{\mu})
-
-\inferrule{
-	\tentailsX{\delta}\\
-	\ientails{\iMkCtx{\fCtx}{\tvarCtx}{\itvarCtx}, \ivar{f}}{\mu}
-}{
-	\ientailsX{\ifix}{\ivar{f}}{\delta}{\mu}
-}~(\text{fix-form}_\mu)
-
-\inferrule{
-	\ientailsX{\mu_1}\\
-	\ientailsX{\mu_2}
-}{
-	\ientailsX{\iapp}{\mu_1}{\mu_2}
-}~(\text{app-form}_\mu)
-
-\inferrule{
-	\ientailsX{\mu_1}\\
-	\ientailsX{\mu_2}\\
-	\ientailsX{\mu_3}\\
-	\ientailsX{\mu_4}
-}{
-	\ientailsX{\iIfEq}{\mu_1}{\mu_2}{\mu_3}{\mu_4}
-}~(\text{cond-form}_\mu)
-
-\inferrule{ }{\ientailsX{\iintlit}}~(\text{int-form}_\mu)
-
-\inferrule{
-	\ientailsX{\mu_1}\\
-	\ientailsX{\mu_2}
-}{
-	\ientailsX{\iop{\mu_1}{\mu_2}}
-}~(\text{+-form}_\mu)
-
-\inferrule{
-	\ientailsX{\mu_1}\\
-	\ientailsX{\mu_2}
-}{
-	\ientailsX{\ipair{\mu_1}{\mu_2}}
-}~(\text{$\times$-form}_\mu)
-
-\inferrule{
-	\ientailsX{\mu}
-}{
-	\ientailsX{\ifst{\mu}}
-}~(\text{fst-form}_\mu)
-
-\inferrule{
-	\ientailsX{\mu}
-}{
-	\ientailsX{\isnd{\mu}}
-}~(\text{snd-form}_\mu)
-
-\inferrule{
-	\tentailsXt{\psi}{\kITerm}
-}{
-	\ientailsX{\iup{\psi}}
-}~(\text{$\uparrow$-form}_\mu)
-\end{mathpar}
-$\fbox{\inferrule{}{\tentailsX{\delta}}}$
-\begin{mathpar}
-\small
-\inferrule{
-	\tentailsX{\delta_1}\\
-	\tentailsX{\delta_2}
-}{
-	\tentailsX{\darrow{\delta_1}{\delta_2}}
-}~(\text{$\rightarrow$-form}_\delta)
-
-\inferrule{ }{\tentailsX{\dint}}~(\text{int-form}_\delta)
-
-\inferrule{
-	\tentailsX{\delta_1}\\
-	\tentailsX{\delta_2}
-}{
-	\tentailsX{\dpair{\delta_1}{\delta_2}}
-}~(\text{$\times$-form}_\delta)
-
-\inferrule{
-	\tentailsXt{\psi}{\kIType}
-}{
-	\tentailsX{\dup{\psi}}
-}~(\text{$\uparrow$-form}_\delta)
-\end{mathpar}
-\caption{Kind and term formation rules}
-\end{figure*}
-
-\begin{figure*}
-%  \renewcommand\thefigure{4}
-% % \begin{textblock}{20}(0,0)
-%  \newpage
-$\fbox{$\tevalX{\psi}{\psi'}$}$~
-$\fbox{$\tevalX{\theta}{\theta'}$}$
-$\fbox{$\tevalX{\phi}{\phi'}$}$
-~~~~$\famEvalCtx ::= \cdot \pipe \famEvalCtx,\tfamSpecStd$
-~~~~~~~~$\errCtx ::= \noprob \pipe \prob \pipe \errCtx, \errCtx$
-\begin{mathpar}
-\small
-\inferrule{ }{
-	\tevalsX{\tunit}{\tunit}{\noprob}
-}~(\text{unit}_\beta)
-
-\inferrule{
-	\tevalsX{\psi_1}{\psi_1'}{\errCtx_{1}}\\
-	\tevalsX{\psi_2}{\psi_2'}{\errCtx_{2}}
-}{
-	\tevalsX{\tpair{\psi_1}{\psi_2}}{\tpair{\psi_1'}{\psi_2'}}{\errCtx_{1},\errCtx_{2}}
-}~(\text{pair}_\beta)
-
-\inferrule{
-	\tevalX{\psi}{\tpair{\psi_1}{\psi_2}}
-}{
-	\tevalX{\tfst{\psi}}{\psi_1}
-}~(\text{fst}_\beta)
-
-\inferrule{
-	\tevalX{\psi}{\tpair{\psi_1}{\psi_2}}
-}{
-	\tevalX{\tsnd{\psi}}{\psi_2}
-}~(\text{snd}_\beta)
-
-\inferrule{ }{
-	\tevalsX{\tlam{\splatTwo{\tvar{t}}{{:}\kappa}{1}{n}}{\psi}}{\tlam{\splatTwo{\tvar{t}}{{:}\kappa}{1}{n}}{\psi}}{\noprob}
-}~(\lambda_\beta)
-
-\inferrule{
-	\tevalsX{\psi_0}{\tlam{\splatTwo{\tvar{t}}{{:}\kappa}{1}{n}}{\psi}}{\errCtx_{0}}\\
-	\tevalsX{\psi_1}{\psi_1'}{\errCtx_{1}}\\
-	\cdots\\
-	\tevalsX{\psi_n}{\psi_n'}{\errCtx_{n}}\\
-	\tevalsX{\substn{\splatTwo{\psi'}{{/}t}{1}{n}}{\psi}}{\psi'}{\errCtx}
-}{
-	\tevalsX{\tapp{\psi_0}{\splat{\psi}{1}{n}}}{\psi'}{\errCtx_{0},\errCtx_{1},\ldots,\errCtx_{n},\errCtx}
-}~(\text{app}_\beta)
-
-\inferrule{
-	\tevalsX{\psi_0}{\psi_0'}{\errCtx_{0}}\\
-	\tevalsX{\psi_1}{\psi_0'}{\errCtx_{1}}\\
-	\tevalsX{\psi_2}{\psi_2'}{\errCtx_{2}}
-}{
-	\tevalsX{\tifeq{\psi_0}{\psi_1}{\psi_2}{\psi_3}}{\psi_2'}{\errCtx_{0},\errCtx_{1},\errCtx_{2}}
-}~(\text{if}_\beta^1)
-
-\inferrule{
-	\tevalsX{\psi_0}{\psi_0'}{\errCtx_{0}}\\
-	\tevalsX{\psi_1}{\psi_1'}{\errCtx_{1}}\\
-	\psi_0' \neq \psi_1'\\
-	\tevalsX{\psi_3}{\psi_3'}{\errCtx_{3}}
-}{
-	\tevalsX{\tifeq{\psi_0}{\psi_1}{\psi_2}{\psi_3}}{\psi_3'}{\errCtx_{0},\errCtx_{1},\errCtx_{3}}
-}~(\text{if}_\beta^2)
-
-\inferrule{
-	\tevalsX{\tfamSpecStd}{\phi}{\errCtx_0}\\
-	\tevalsX{\subst{\phi}{\fvar{fam}}{\psi}}{\psi'}{\errCtx_1}
-}{
-	\tevalsX{\tfamStd}{\psi'}{\errCtx_0, \errCtx_1}
-}~(\text{fambind}_\beta)
-
-\inferrule{
-	\tevalsX{\psirep}{\psirep'}{\errCtx_0}\\
-	\tevalsX{\theta}{\theta'}{\errCtx_1}
-}{
-	\tevalsX{\tfamSpecStd}{\tfamSpec{fam}{\kappaidx}{\psirep'}{\theta'}{\Theta}}{\errCtx_0, \errCtx_1}
-}~(\text{family}_\beta)
-
-\inferrule{
-	\tevalsX{\psitype}{\ttype{\psiidx}{\phi'}}{\errCtx_0}\\
-	\tevalsX{\phi}{\phi'}{\errCtx_1}\\
-	\tevalsX{\tapp{\psi_0}{\ttype{\psiidx}{\phi'}}}{\psi_0'}{\errCtx_2}
-}{
-	\tevalsX{
-		\tfamcase{\psitype}{\phi}{\psi_0}{\psi_1}
-	}{
-		\psi_0'
-	}{\errCtx_0, \errCtx_1, \errCtx_2}
-}~(\text{famcase-T}_\beta^1)
-
-\inferrule{
-	\tevalsX{\psitype}{\ttype{\psiidx}{\phi'}}{\errCtx_0}\\
-	\tevalsX{\phi}{\phi''}{\errCtx_1}\\
-	\phi' \neq \phi''\\
-	\tevalsX{\psi_1}{\psi_1'}{\errCtx_2}
-}{
-	\tevalsX{
-		\tfamcase{\psitype}{\phi}{\psi_0}{\psi_1}
-	}{
-		\psi_1'
-	}{\errCtx_0, \errCtx_1, \errCtx_2}
-}~(\text{famcase-T}_\beta^2)
-
-\inferrule{
-	\tevalsX{\psiden}{\tden{\psiiterm}{\ttype{\psiidx}{\phi'}}}{\errCtx_0}\\
-	\tevalsX{\phi}{\phi'}{\errCtx_1}\\
-	\tevalsX{\tapp{\psi_0}{\tden{\psiiterm}{\ttype{\psiidx}{\phi'}}}}{\psi_0'}{\errCtx_2}
-}{
-	\tevalsX{
-		\tfamcase{\psiden}{\phi}{\psi_0}{\psi_1}
-	}{
-		\psi_0'
-	}{\errCtx_0, \errCtx_1, \errCtx_2}
-}~(\text{famcase-D}_\beta^1)
-
-\inferrule{
-	\tevalsX{\psiden}{\tden{\psiiterm}{\ttype{\psiidx}{\phi'}}}{\errCtx_0}\\
-	\tevalsX{\phi}{\phi''}{\errCtx_1}\\
-	\phi' \neq \phi''\\
-	\tevalsX{\psi_1}{\psi_1'}{\errCtx_2}
-}{
-	\tevalsX{
-		\tfamcase{\psiden}{\phi}{\psi_0}{\psi_1}
-	}{
-		\psi_1'
-	}{\errCtx_0, \errCtx_1, \errCtx_2}
-}~(\text{famcase-D}_\beta^2)
-
-\inferrule{
-	\tevalsX{\psiidx}{\psiidx'}{\errCtx}
-}{
-	\tevalsX{\ttypestd}{
-		\ttype{\psiidx'}{fam}
-	}{\errCtx}
-}
-
-\inferrule{
-	\tevalsX{\psitype}{\ttypestd}{\errCtx}
-}{
-	\tevalsX{\tidx{\psitype}}{\psiidx}{\errCtx}
-}
-
-\inferrule{
-	\tevalsX{\psitype}{\ttypestd}{\errCtx_0}\\
-	\tfamSpecStd \in \famEvalCtx\\
-	\tevalsX{\tapp{\psirep}{\psiidx}}{\psi'}{\errCtx_1}
-}{
-	\tevalsX{\trepof{\psitype}}{\psi'}{\errCtx_0, \errCtx_1}
-}
-
-\inferrule{
-	\tevalsX{\psi}{\psi'}{\errCtx}
-}{
-	\tevalsX{\tconst{\psi}}{\tconst{\psi'}}{\errCtx}
-}
-
-\inferrule{
-	\tevalsX{\psi}{\psi'}{\errCtx}
-}{
-	\tevalsX{\tOp{\psi}}{\tOp{\psi'}}{\errCtx}
-}
-
-\inferrule{
-	\tevalsX{\psiiterm}{\psiiterm'}{\errCtx_0}\\
-	\tevalsX{\psitype}{\psitype'}{\errCtx_1}
-}{
-	\tevalsX{\tden{\psiiterm}{\psitype}}{\tden{\psiiterm'}{\psitype'}}{\errCtx_0, \errCtx_1}
-}
-
-\inferrule{
-	\tevalsX{\psiden}{\tden{\psiiterm}{\psitype}}{\errCtx}
-}{
-	\tevalsX{\tvalof{\psiden}}{\psiiterm}{\errCtx}
-}
-
-\inferrule{
-	\tevalsX{\psiden}{\tden{\psiiterm}{\psitype}}{\errCtx}
-}{
-	\tevalsX{\ttypeof{\psiden}}{\psitype}{\errCtx}
-}
-
-\inferrule{ }{
-	\tevalsX{\terr}{\terr}{\noprob}
-}
-
-\inferrule{
-	\tevalsX{\mu}{\mu'}{\errCtx}
-}{
-	\tevalsX{\titerm{\mu}}{\titerm{\mu'}}{\errCtx}
-}
-
-\inferrule{
-	\tevalsX{\delta}{\delta'}{\errCtx}
-}{
-	\tevalsX{\titype{\delta}}{\titype{\delta'}}{\errCtx}
-}
-
-\inferrule{
-	\tevale{e}{e'}
-}{
-	\tevalX{\tprog{e}}{\tprog{e'}}
-}
-
-\inferrule{ }{
-	\tevalsX{\topsempty}{\topsempty}{\noprob}
-}
-
-\inferrule{
-	\tevalsX{\theta}{\theta'}{\errCtx_0}\\
-	\tevalsX{\psi}{\psi'}{\errCtx_1}
-}{
-	\tevalsX{
-		\topp{\theta}{\tops{op}{\psi}}
-	}{
-		\topp{\theta'}{\tops{op}{\psi'}}
-	}{\errCtx_0, \errCtx_1}
-}
-\end{mathpar}
-$\fbox{$\tevale{e}{e'}$}$
-\begin{mathpar}
-\small
-\inferrule{
-	\tevalsX{\psi}{\psi'}{\errCtx}
-}{
-	\tevalsX{\elam{\evar{x}}{\psi}{e}}
-	{\elam{\evar{x}}{\psi'}{e}}{\errCtx}
-}
-
-\inferrule{
-	\tevalsX{e_1}{e_1'}{\errCtx_1}\\
-	\tevalsX{e_2}{e_2'}{\errCtx_2}
-}{
-	\tevalsX{\eapp{e_1}{e_2}}{\eapp{e_1'}{e_2'}}{\errCtx_1, \errCtx_2}
-}
-
-\inferrule{
-	\tevalsX{\psitype}{\psitype'}{\errCtx}\\
-	\tevalsX{\psi_1}{\psi_1'}{\errCtx_1^1}\\
-	\cdots\\
-	\tevalsX{\psi_m}{\psi_m'}{\errCtx_m^1}\\
-	\tevalsX{e_1}{e_1'}{\errCtx_1^2}\\
-	\cdots\\
-	\tevalsX{e_n}{e_n'}{\errCtx_n^2}
-}{
-	\tevalsX{\eop{\psitype}{op}{\splat{\psi}{1}{m}}{\splat{e}{1}{n}}}
-	{\eop{\psitype'}{op}{\splat{\psi'}{1}{m}}{\splat{e'}{1}{n}}}
-	{\errCtx, \errCtx_1^1, \ldots, \errCtx_m^1, \errCtx_1^2, \ldots, \errCtx_n^2}
-}
-\end{mathpar}
-$\fbox{$\tevalm{\mu}{\mu'}$}$
-\begin{mathpar}
-\small
-\inferrule{
-	\tevalsX{\delta}{\delta'}{\errCtx}
-}{
-	\tevalsX{\ilam{\ivar{u}}{\delta}{\mu}}{\ilam{\ivar{u}}{\delta'}{\mu}}{\errCtx}
-}
-
-\inferrule{
-	\tevalsX{\delta}{\delta'}{\errCtx}
-}{
-	\tevalsX{\ifix{\ivar{f}}{\delta}{\mu}}{\ifix{\ivar{f}}{\delta'}{\mu}}{\errCtx}
-}
-
-\inferrule{
-	\tevalsX{\mu_1}{\mu_1'}{\errCtx_1}\\
-	\tevalsX{\mu_2}{\mu_2'}{\errCtx_2}
-}{
-	\tevalsX{\iapp{\mu_1}{\mu_2}}{\iapp{\mu_1'}{\mu_2'}}{\errCtx_1, \errCtx_2}
-}
-
-\inferrule{
-	\tevalsX{\mu_0}{\mu_0'}{\errCtx_0}\\
-	\tevalsX{\mu_1}{\mu_1'}{\errCtx_1}\\
-	\tevalsX{\mu_2}{\mu_2'}{\errCtx_2}\\
-	\tevalsX{\mu_3}{\mu_3'}{\errCtx_3}
-}{
-	\tevalsX{\iIfEq{\mu_0}{\mu_1}{\mu_2}{\mu_3}}{\iIfEq{\mu_0'}{\mu_1'}{\mu_2'}{\mu_3'}}{\errCtx_0, \errCtx_1, \errCtx_2, \errCtx_3}
-}
-
-\inferrule{ }{
-	\tevalsX{\iintlit}{\iintlit}{\noprob}
-}
-
-\inferrule{
-	\tevalsX{\mu_1}{\mu_1'}{\errCtx_1}\\
-	\tevalsX{\mu_2}{\mu_2'}{\errCtx_2}
-}{
-	\tevalsX{\iop{\mu_1}{\mu_2}}{\iop{\mu_1'}{\mu_2'}}{\errCtx_1, \errCtx_2}
-}
-
-\inferrule{
-	\tevalsX{\mu_1}{\mu_1'}{\errCtx_1}\\
-	\tevalsX{\mu_2}{\mu_2'}{\errCtx_2}
-}{
-	\tevalsX{\ipair{\mu_1}{\mu_2}}{\ipair{\mu_1'}{\mu_2'}}{\errCtx_1, \errCtx_2}
-}
-
-\inferrule{
-	\tevalsX{\mu}{\mu'}{\errCtx}
-}{
-	\tevalsX{\ifst{\mu}}{\ifst{\mu'}}{\errCtx}
-}
-
-\inferrule{
-	\tevalsX{\mu}{\mu'}{\errCtx}
-}{
-	\tevalsX{\isnd{\mu}}{\isnd{\mu'}}{\errCtx}
-}
-
-\inferrule{
-	\tevalsX{\psi}{\titerm{\mu}}{\errCtx}
-}{
-	\tevalsX{\iup{\psi}}{\mu}{\errCtx}
-}
-\end{mathpar}
-$\fbox{$\tevalm{\delta}{\delta'}$}$
-\begin{mathpar}
-\small
-\inferrule{
-	\tevalsX{\delta_1}{\delta_1'}{\errCtx_1}\\
-	\tevalsX{\delta_2}{\delta_2'}{\errCtx_2}
-}{
-	\tevalsX{\darrow{\delta_1}{\delta_2}}{\darrow{\delta_1'}{\delta_2'}}{\errCtx_1, \errCtx_2}
-}
-
-\inferrule{ }{
-	\tevalsX{\dint}{\dint}{\noprob}
-}
-
-\inferrule{
-	\tevalsX{\delta_1}{\delta_1'}{\errCtx_1}\\
-	\tevalsX{\delta_2}{\delta_2'}{\errCtx_2}
-}{
-	\tevalsX{\dpair{\delta_1}{\delta_2}}{\dpair{\delta_1'}{\delta_2'}}{\errCtx_1, \errCtx_2}
-}
-
-\inferrule{
-	\tevalsX{\psi}{\titype{\delta}}{\errCtx}
-}{
-	\tevalsX{\dup{\psi}}{\delta}{\errCtx}
-}
-\end{mathpar}
-%
-%\inferrule{
-%	\tevalX{\psi}{\ttype{\psiidx}{\kappa}{\psirep}}
-%}{
-%	\tevalX{\tidx{\psi}}{\psiidx}
-%}
-%
-%\inferrule{
-%	\tevalX{\psi}{\ttype{\psiidx}{\kappa}{\psirep}}
-%}{
-%	\tevalX{\trepof{\psi}}{\psirep}
-%}
-%
-%\inferrule{
-%	\tevalsX{\psi}{\ttype{\psiidx}{\kappa}{\psirep}}{\errCtx_{0}}\\
-%	\tevalsX{\tapp{\psi_{0}}{\ttype{\psiidx}{\kappa}{\psirep}}}{\psi'}{\errCtx_{1}}
-%}{
-%	\tevalsX{\tfamcase{\psi}{\kappa}{\psi_{0}}{\psi_{1}}}{\psi'}{\errCtx_{0},\errCtx_{1}}
-%}
-%
-%\inferrule{
-%	\tevalsX{\psi}{\ttype{\psiidx}{\kappa'}{\psirep}}{\errCtx_{0}}\\
-%	\kappa \neq \kappa'\\
-%	\tevalsX{\psi_{1}}{\psi'}{\errCtx_{1}}
-%}{
-%	\tevalsX{\tfamcase{\psi}{\kappa}{\psi_{0}}{\psi_{1}}}{\psi'}{\errCtx_{0},\errCtx_{1}}
-%}
-%
-%\inferrule{
-%	\tevalX{\psi}{\psi'}
-%}{
-%	\tevalX{\tconst{\psi}}{\tconst{\psi'}}
-%}
-%
-%\inferrule{
-%	\tevalX{\psi}{\psi'}
-%}{
-%	\tevalX{\tOp{\psi}}{\tOp{\psi'}}
-%}
-%
-%\inferrule{
-%	\tevalms{\mu}{\mu'}{\errCtx_{0}}\\
-%	\tevalsX{\psitype}{\ttype{\psiidx}{\kappa}{\psirep}}{\errCtx_{1}}\\
-%	\mtentails{\cdot}{\mtof{\mu'}{\psirep}}
-%}{
-%	\tevalsX{\tden{\mu}{\psitype}}{\tden{\mu'}{\psitype'}}{\errCtx_{0},\errCtx_{1}}
-%}
-%
-%\inferrule{
-%	\tevalms{\mu}{\mu'}{\errCtx_{0}}\\
-%	\tevalsX{\psitype}{\ttype{\psiidx}{\kappa}{\psirep}}{\errCtx_{1}}\\
-%	\cdot \nvdash \mtof{\mu'}{\psirep}
-%}{
-%	\tevalsX{\tden{\mu}{\psitype}}{\tden{\mu'}{\psitype'}}{\errCtx_{0},\errCtx_{1},\prob}
-%}
-%
-%\inferrule{
-%	\tevalX{\psi}{\tden{\mu}{\psitype}}
-%}{
-%	\tevalX{\ttypeof{\psi}}{\psitype}
-%}
-%
-%\inferrule{ }{
-%	\tevalsX{\terr}{\terr}{\noprob}
-%}
-%
-%\inferrule{
-%	\tevalsX{\psi}{\tden{\mu}{\ttype{\psiidx}{\kappa}{\psirep}}}{\errCtx_{0}}\\
-%	\tevalsX{\tapp{\psi_{0}}{\tden{\mu}{\ttype{\psiidx}{\kappa}{\psirep}}}}{\psi'}{\errCtx_{1}}
-%}{
-%	\tevalsX{\ttidxcase{\psi}{\kappa}{\psi_{0}}{\psi_{1}}}{\psi'}{\errCtx_{0},\errCtx_{1}}
-%}
-%
-%\inferrule{
-%	\tevalsX{\psi}{\terr}{\errCtx_{0}}\\
-%	\tevalsX{\psi_{1}}{\psi'}{\errCtx_{1}}
-%}{
-%	\tevalsX{\ttidxcase{\psi}{\kappa}{\psi_{0}}{\psi_{1}}}{\psi'}{\errCtx_{0},\errCtx_{1}}
-%}
-%
-%\inferrule{
-%	\tevalsX{\psi}{\tden{\mu}{\ttype{\psiidx}{\kappa'}{\psirep}}}{\errCtx_{0}}\\
-%	\kappa \neq \kappa'\\
-%	\tevalsX{\psi_{1}}{\psi'}{\errCtx_{1}}
-%}{
-%	\tevalsX{\ttidxcase{\psi}{\kappa}{\psi_{0}}{\psi_{1}}}{\psi'}{\errCtx_{0},\errCtx_{1}}
-%}
-%
-%\inferrule{
-%	\tevalsX{\psi_1}{\psi_1'}{\errCtx_{0}}\\
-%	\tevalsX{\psi_2}{\psi_2'}{\errCtx_{1}}
-%}{
-%	\tevalsX{\tILtarrow{\psi_1}{\psi_2}}{\tILtarrow{\psi_1'}{\psi_2'}}{\errCtx_{0},\errCtx_{1}}
-%}
-%
-%\inferrule{
-%	\tevalsX{\psi_1}{\psi_1'}{\errCtx_{0}}\\
-%	\tevalsX{\psi_2}{\psi_2'}{\errCtx_{1}}
-%}{
-%	\tevalsX{\tILtprod{\psi_1}{\psi_2}}{\tILtprod{\psi_1'}{\psi_2'}}{\errCtx_{0},\errCtx_{1}}
-%}
-%
-%\inferrule{ }{
-%	\tevalsX{\tILtint}{\tILtint}{\noprob}
-%}
-%
-%\inferrule{
-%	\tevalsX{\psi}{\tILtarrow{\psi_1}{\psi_2}}{\errCtx_{0}}\\
-%	\tevalsX{\tapp{\psiarrow}{\psi_1, \psi_2}}{\psi'}{\errCtx_{1}}
-%}{
-%	\tevalsX{\tILtcase{\psi}{\psiarrow}{\psiprod}{\psiint}}{\psi'}{\errCtx_{0},\errCtx_{1}}
-%}
-%
-%\inferrule{
-%	\tevalsX{\psi}{\tILtprod{\psi_1}{\psi_2}}{\errCtx_{0}}\\
-%	\tevalsX{\tapp{\psiprod}{\psi_1, \psi_2}}{\psi'}{\errCtx_{1}}
-%}{
-%	\tevalsX{\tILtcase{\psi}{\psiarrow}{\psiprod}{\psiint}}{\psi'}{\errCtx_{0},\errCtx_{1}}
-%}
-%
-%\inferrule{
-%	\tevalsX{\psi}{\tILtint}{\errCtx_{0}}\\
-%	\tevalsX{\psiint}{\psiint'}{\errCtx_{1}}
-%}{
-%	\tevalsX{\tILtcase{\psi}{\psiarrow}{\psiprod}{\psiint}}{\psiint'}{\errCtx_{0},\errCtx_{1}}
-%}
-%
-%\inferrule{ }{
-%	\tevalms{\ivar{u}}{\ivar{u}}{\noprob}
-%}
-%
-%\inferrule{
-%	\tevalX{\psiden}{\tden{\mu}{\psitype}}
-%}{
-%	\tevalm{\mvalof{\psiden}}{\mu}
-%}
-%
-%\inferrule{
-%	\tevalsX{\psi}{\psi'}{\errCtx_{0}}\\
-%	\tevalms{\mu}{\mu'}{\errCtx_{1}}
-%}{
-%	\tevalms{\ilam{\ivar{u}}{\psi}{\mu}}{\ilam{\ivar{u}}{\psi'}{\mu'}}{\errCtx_{0},\errCtx_{1}}
-%}
-%
-%\inferrule{
-%	\tevalsX{\psi}{\psi'}{\errCtx_{0}}\\
-%	\tevalms{\mu}{\mu'}{\errCtx_{1}}
-%}{
-%	\tevalms{\ifix{\ivar{f}}{\psi}{\mu}}{\ifix{\ivar{f}}{\psi'}{\mu'}}{\errCtx_{0},\errCtx_{1}}
-%}
-%
-%\inferrule{
-%	\tevalms{\mu_1}{\mu_1'}{\errCtx_{0}}\\
-%	\tevalms{\mu_2}{\mu_2'}{\errCtx_{1}}
-%}{
-%	\tevalms{\iapp{\mu_1}{\mu_2}}{\iapp{\mu_1'}{\mu_2'}}{\errCtx_{0},\errCtx_{1}}
-%}
-%
-%\inferrule{ }{
-%	\tevalms{\iintlit}{\iintlit}{\noprob}
-%}
-%
-%\inferrule{
-%	\tevalms{\mu_1}{\mu_1'}{\errCtx_{0}}\\
-%	\tevalms{\mu_2}{\mu_2'}{\errCtx_{1}}
-%}{
-%	\tevalms{\iop{\mu_1}{\mu_2}}{\iop{\mu_1'}{\mu_2'}}{\errCtx_{0},\errCtx_{1}}
-%}
-%
-%\inferrule{
-%	\tevalms{\mu_1}{\mu_1'}{\errCtx_{0}}\\
-%	\tevalms{\mu_2}{\mu_2'}{\errCtx_{1}}
-%}{
-%	\tevalms{\ipair{\mu_1}{\mu_2}}{\ipair{\mu_1'}{\mu_2'}}{\errCtx_{0},\errCtx_{1}}
-%}
-%
-%\inferrule{
-%	\tevalm{\mu}{\mu'}
-%}{
-%	\tevalm{\ifst{\mu}}{\ifst{\mu'}}
-%}
-%
-%\inferrule{
-%	\tevalm{\mu}{\mu'}
-%}{
-%	\tevalm{\isnd{\mu}}{\isnd{\mu'}}
-%}
-%
-%\inferrule{
-%	\tevalms{\mu_0}{\mu_0'}{\errCtx_{0}}\\
-%	\tevalms{\mu_1}{\mu_1'}{\errCtx_{1}}\\
-%	\tevalms{\mu_2}{\mu_2'}{\errCtx_{2}}\\
-%	\tevalms{\mu_3}{\mu_3'}{\errCtx_{3}}
-%}{
-%	\tevalms{\iIfEq{\mu_0}{\mu_1}{\mu_2}{\mu_3}}{\iIfEq{\mu_0'}{\mu_1'}{\mu_2'}{\mu_3'}}{\errCtx_{0},\errCtx_{1},\errCtx_{2},\errCtx_{3}}
-%}
-%
-%\inferrule{ }{
-%	\tevales{\evar{x}}{\evar{x}}{\noprob}
-%}
-%
-%\inferrule{
-%	\tevalsX{\psi}{\psi'}{\errCtx_{0}}\\
-%	\tevales{e}{e'}{\errCtx_{1}}
-%}{
-%	\tevales{\elam{\evar{x}}{\psi}{e}}{\elam{\evar{x}}{\psi'}{e'}}{\errCtx_{0},\errCtx_{1}}
-%}
-%
-%\inferrule{
-%	\tevales{e_1}{e_1'}{\errCtx_{0}}\\
-%	\tevales{e_2}{e_2'}{\errCtx_{1}}
-%}{
-%	\tevales{\eapp{e_1}{e_2}}{\eapp{e_1'}{e_2'}}{\errCtx_{0},\errCtx_{1}}
-%}
-%
-%\inferrule{
-%	\tevalX{\psi}{\psi'}
-%}{
-%	\tevale{\econst{\psi}}{\econst{\psi'}}
-%}
-%
-%\inferrule{
-%	\tevalsX{\psi}{\psi'}{\errCtx_{0}}\\
-%	\tevales{e_1}{e_1'}{\errCtx_{1}}\\
-%	\cdots\\
-%	\tevales{e_n}{e_n'}{\errCtx_{n}}
-%}{
-%	\tevales{\eop{\psi}{\splat{e}{1}{n}}}{\eop{\psi'}{\splat{e'}{1}{n}}}{\errCtx_{0},\errCtx_{1},\ldots,\errCtx_{n}}
-%}
-\caption{Big-step operational semantics for type-level terms. The error context $\Sigma$ is used to detect representational consistency errors.}
-%%\end{textblock}
-\end{figure*}
-%\begin{figure*}
-%\renewcommand\thefigure{5}
-%$\fbox{$\translates{\eCtx}{e}{\mu}{\psiidx}{\kappa}{\psirep}$}$
-%~~~~$\eCtx ::= \cdot \pipe \extendeCtx{x}{\mu}{\psiidx}{\kappa}{\psirep}$
-%\begin{mathpar}
-%\inferrule{
-%	\translatesTo{\evar{x}}{\mu}{\psiidx}{\kappa}{\psirep} \in \eCtx
-%}{
-%	\translatesX{\evar{x}}{\mu}{\psiidx}{\kappa}{\psirep}
-%}
-%
-%\inferrule{ }{
-%	\translatesX{
-%		\econst{\tconst{\tden{\mu}{\ttype{\psiidx}{\kappa}{\psirep}}}}
-%	}{
-%		\mu
-%	}{
-%		\psiidx
-%	}{
-%		\kappa
-%	}{
-%		\psirep
-%	}
-%}
-%
-%\inferrule{
-%	\translates{\extendeCtx{x}{\ivar{x}}{\psiidx}{\kappa}{\psirep}}{e}{\mu'}{\psiidx'}{\kappa'}{\psirep'}
-%}{
-%	\translatesX{
-%		\elam{\evar{x}}{\ttype{\psiidx}{\kappa}{\psirep}}{e}
-%	}{
-%		\ilam{\ivar{x}}{\psirep}{\mu'}
-%	}{
-%		\tpair{\ttype{\psiidx}{\kappa}{\psirep}}{\ttype{\psiidx'}{\kappa'}{\psirep'}}
-%	}{
-%		\kpair{\kType{\kappa}}{\kType{\kappa'}}
-%	}{
-%		\tILtarrow{\psirep}{\psirep'}
-%	}
-%}
-%
-%\inferrule{
-%	\translatesX{e_1}{\mu_1}{\psiidxn{1}}{\kappa_1}{\psirepn{1}}~~~~
-%	\cdots~~~~
-%	\translatesX{e_n}{\mu_n}{\psiidxn{n}}{\kappa_n}{\psirepn{n}}\\
-%	\tevalsX{\tapp{\psi}{
-%		\tden{\mu_1}{\ttype{\psiidxn{1}}{\kappa_1}{\psirepn{1}}},
-%		\ldots,
-%		\tden{\mu_n}{\ttype{\psiidxn{n}}{\kappa_n}{\psirepn{n}}}
-%	}}{\tden{\mu}{\ttype{\psiidx}{\kappa}{\psirep}}}{\noprob}
-%}{
-%	\translatesX{
-%		\eop{\tOp{\psi}}{\splat{e}{1}{n}}
-%	}{
-%		\mu
-%	}{
-%		\psiidx
-%	}{
-%		\kappa
-%	}{
-%		\psirep
-%	}
-%}
-%
-%\end{mathpar}
-%\caption{Verification and translation rules for $\lamA$.}
-%\end{figure*}
-%\begin{figure*}
-%  \renewcommand\thefigure{6}
-%$\fbox{$\mtentailsXt{\mu}{\psi}$}$
-%~~~~$\iCtx ::= \cdot \pipe \iCtx, \mtof{\ivar{u}}{\psi}$
-%\begin{mathpar}
-%\inferrule{
-%	\mtof{\ivar{u}}{\psi} \in \ivarCtx
-%}{
-%	\mtentailsXt{\ivar{u}}{\psi}
-%}
-%
-%\inferrule{
-%	\mtentails{\iCtx, \mtof{\ivar{u}}{\psi}}{\mtof{\mu}{\psi'}}
-%}{
-%	\mtentailsXt{\ilam{\ivar{u}}{\psi}{\mu}}{\psi'}
-%}
-%
-%\inferrule{
-%	\mtentails{\iCtx, \mtof{\ivar{f}}{\psi}}{\mtof{\mu}{\psi}}
-%}{
-%	\mtentailsXt{\ifix{\ivar{f}}{\psi}{\mu}}{\psi}
-%}
-%
-%\inferrule{
-%	\mtentailsXt{\mu_1}{\tILtarrow{\psi_1}{\psi_2}}\\
-%	\mtentailsXt{\mu_2}{\psi_1}
-%}{
-%	\mtentailsXt{\iapp{\mu_1}{\mu_2}}{\psi_2}
-%}
-%
-%\inferrule{ }{
-%	\mtentailsXt{\iintlit}{\tILtint}
-%}
-%
-%\inferrule{
-%	\mtentailsXt{\mu_1}{\tILtint}\\
-%	\mtentailsXt{\mu_2}{\tILtint}
-%}{
-%	\mtentailsXt{\iop{\mu_1}{\mu_2}}{\tILtint}
-%}
-%
-%\inferrule{
-%	\mtentailsXt{\mu_1}{\psi_1}\\
-%	\mtentailsXt{\mu_2}{\psi_2}
-%}{
-%	\mtentailsXt{\ipair{\mu_1}{\mu_2}}{\tILtprod{\psi_1}{\psi_2}}
-%}
-%
-%\inferrule{
-%	\mtentailsXt{\mu}{\tILtprod{\psi_1}{\psi_2}}
-%}{
-%	\mtentailsXt{\ifst{\mu}}{\psi_1}
-%}
-%
-%\inferrule{
-%	\mtentailsXt{\mu}{\tILtprod{\psi_1}{\psi_2}}
-%}{
-%	\mtentailsXt{\isnd{\mu}}{\psi_2}
-%}
-%
-%\inferrule{
-%	\mtentailsXt{\mu_0}{\psi_1}\\
-%	\mtentailsXt{\mu_1}{\psi_1}\\
-%	\mtentailsXt{\mu_2}{\psi_2}\\
-%	\mtentailsXt{\mu_3}{\psi_2}
-%}{
-%	\mtentailsXt{\iIfEq{\mu_0}{\mu_1}{\mu_2}{\mu_3}}{\psi_2}
-%}
-%\end{mathpar}
-%\caption{The internal type system of $\lamA$.}
-%\end{figure*}
-
-%Let us now examine the definition of $\lamA$ in more detail.
-The grammar of $\lamA$ is given in Figure 2.
-Expressions in $\lamA$ can take one of only five forms, the first three of which correspond to the simply-typed lambda calculus and the remaining two to abstract constants and operator invocations, respectively.
-%
-%The type-level specification language is based on System F \cite{tapl} equipped with several additional primitives that we will describe below.
-To emphasize the expanded role for type-level terms, we use the letter $\psi$ instead of the more conventional $\tau$.
-It is constrained by a kind system, the rules of which are given in Figure 4.
-A big-step semantics for the type-level language is given in Figure 5.
-As we will state more formally below, well-kinded terms of the type-level language strongly normalize.
-%
-%Expressions are given meaning in a denotational style in terms of a typed internal language according to the verification and translation rules in Figure 6.
-The type system of the internal language is given in Figure 7.
-Unlike the type-level language, the internal language contains a ${\sf fix}$ operator, and is thus Turing-complete (corresponding to a variant of Plotkin's PCF \cite{tapl}.)
-%
-%\subsubsection{Central Compilation Judgement}
-%The central compilation judgement of Figure 3 takes a type-level term that reduces to a program and produces a typed denotation for the program if and only if it is valid with respect to the kind system of $\lamA$ {\it and} the type system introduced by the user, the key feature of this mechanism.
-It consists of three steps:
-%\begin{enumerate}
-%\item \textbf{Kind checking}, where basic correctness properties of the user specification as well as uses of type-level terms in expressions are checked.
-%\item \textbf{Specification Normalization}, which reduces the specification to normal form by internalizing relevant portions of it directly into the expressions of the program.
-%\item \textbf{Verification and Translation}, where operator definitions are invoked to assign final denotations to expressions if possible, or derive that there is an $\terr$ if not.
-%\end{enumerate}
-%The rules given in Figures 4-7 are sufficient to define the central compilation judgement. We examine its theoretical properties in Section 2.4.
-%
-%\subsubsection{Type Constructors}
-%
-%An interesting feature of $\lamA$ is that there is just one form of type constructor: 
-%$\ttype{\psiidx}{\kappa}{\psirep}$.
-%All types are identified by their index, $\psiidx$ and an associated internal type, $\psirep$, which controls the representation of values that inhabit the type.
-The principal kind of a type constructor indexed by a value of kind $\kappa$ is $\kType{\kappa}$.
-%
-%
-
-afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf
-
-afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf
-
-afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf
-
-afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf
-
-afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf
-
-afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf
-
-afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf
-
-afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf
-
-afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf
-
-afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf
-
-afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf
-
-afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf
-
-afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf
-
-afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf
-
-afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf
-
-afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf
-
-afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf
-
-afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf
-
-afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf
-
-afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf
-
-afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf
-
-afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf
-
-afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf
-
-afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf
-
-afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf
-
-afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf
-
-%In the example in Figure 1, $\tvar{nat}$ is a type with no (that is, unit) index that we will represent by an integer in the internal language.
-To distinguish it from other types of the same basic form, we need to label it.
-That is the purpose of the \textsf{label} declaration preceding the type definition -- it introduces a new kind of unit labeled $\klabel{nat}$.
-We can then identify the type $\tvar{nat}$ as being the sole type indexed by a value of kind $\klabel{nat}$ and represented by the internal type $\tILtint$.
-It is important to note that although the type $\tvar{nat}$ has kind $\kType{\klabel{nat}}$, it does not uniquely inhabit that kind -- you can construct other types indexed by a $\klabel{nat}$-labeled unit but associated with different internal types, so operator definitions must test for equality as appropriate.
-%
-%\subsubsection{Internal Terms and Types}
-%The internal language, $\mu$, is a variant of Plotkin's PCF with integers, products and a conditional checks for $\beta$-equality.
-The form $\mvalof{\psi}$ is used to compose denotations, which are internal language terms paired with a type constructor (see below).
-In the type-level operational semantics, the $\mvalof{\psi}$ form  always reduces to one of the other forms of the internal language.
-%
-%Internal types are represented at the type-level by the constructors \texttt{arrow}, \texttt{prod} and \texttt{int}, which have kind $\kIType$.
-The internal type system is given in Figure 7.
-The language is equipped with the standard type safety theorem based on preservation and progress lemmas.
-Because expressions are given their semantics ultimately in terms of internal language terms, these global safety theorems are inherited by the language as a whole.
-In a more realistic language, additional desirable global safety restrictions could also be imposed using the internal type system (e.g.
-memory safety).
-%
-%\subsubsection{Function Types}
-%It is worth considering now what type to assign to expression-level functions, e.g.
-$\elam{\evar{x}}{\tvar{nat}}{\eop{\tvar{s}}{\eop{\tvar{s}}{x}}}$.
-Because we have only one form for types, we must encode arrow types using this form.
-The encoding, given formally in Figure 6, encodes the arrow type $\psi_{1} \rightarrow \psi_{2}$, where $\psi_{1,2}$ are types indexed by values of kind $\kappa_{1,2}$, as
-%$$\ttype{\tpair{\psi_{1}}{\psi_{2}}}{\kpair{\kType{\kappa_{1}}}{\kType{\kappa_{2}}}}{\tILtarrow{\trepof{\psi_{1}}}{\trepof{\psi_{2}}}}$$
-%\noindent
-%That is, the arrow type is a type indexed by a pair containing the constituent types and represented by the internal arrow type.
-As can be seen in the compilation rules, expression-level functions translate directly into internal language functions.
-%\subsubsection{Operators}
-%In order to construct and manipulate terms of a newly introduced type, we must provide primitive operators, introduced using the forms \textsf{const} and \textsf{op}.
-The kind associated with operators is $\kOp{n}$, where $n$ is the operator's arity.
-An operator must have fixed arity.
-Constants can be thought of as nullary operators.
-%
-%Constants are given meaning directly using a denotation.
-The constant $\tvar{z}$, for example, has the denotation $\tden{0}{\tvar{nat}}$, reflecting the fact that it is a constant of type $\tvar{nat}$ represented by the value $0$.
-Expressions of the form $\econst{\psi}$ compile to the denotation associated with $\psi$ if $\psi$ is a constant, as can be seen in Figure 6.
-%
-%Operators of higher arity, $n$, are parameterized by a type-level function that takes $n$ denotations (of kind $\kDen$, see below).
-This function must analyze the types associated with these denotations, then return a new denotation composed using the input denotations to give meaning to expressions of the form $\eop{\psi}{\splat{e}{1}{n}}$.
-Note that this implies that operators must have a strictly compositional semantics -- neither expressions nor internal terms can be directly manipulated in the specification language.
-%
-%Because type checking and evaluation of function application is not compositional (requiring extending a context and substitution, respectively), it cannot be simulated by an operator definition in $\lamA$ as stated.
-%
-%\subsubsection{Denotations}
-%Although operator implementations take arguments of kind $\kDen$, a denotation itself does not have principal kind $\kDen$.
-Recall that denotations contain a type, and types have kind $\kType{\kappa}$, where $\kappa$ is the kind of the type's index.
-To support the extraction of a type from a denotation, we must ``remember'' its index kind even when it has been packed into a denotation.
-As a result, denotations have principal kind $\kDenk{\kappa}$.
-%
-%To allow for operators that are polymorphic over a number of types (but not necessarily {\it all} types), we allow for the index kind to be erased as needed.
-To reintroduce the kind, the $\textsf{idxcase}_{D}$ form is available to perform conditional kind analysis during normalization.
-This can be thought of as a simplified form of subkinding with casts.
-Analagous erasure semantics exist for types themselves, with a corresponding {\sf idxcase} form.
-%
-%To ensure that this does not allow for non-terminating terms, we impose a strict positivity restriction on type indices that contain values of kind $\kTypeBlur$.
-This condition, which restricts $\kTypeBlur$ so that it cannot appear to the left of any arrows, is found in many total functional programming languages that feature inductive datatypes, such as Coq \cite{bertot04book} and Agda \cite{norell2007towards}, and we refer the reader to these sources for a detailed discussion.
-%
-%Another important reason for the distinction between $\kDen$ and $\kDenk{\kappa}$ is that there must be some mechanism by which an operator specification indicates a type error.
-In $\lamA$, this corresponds to a special value, $\terr$, which is also of kind $\kDen$.
-However, no type is associated with the denotation $\terr$, so it can never be given the kind $\kDenk{\kappa}$.
-The verification and translation rules in Figure 6 ensure that $\terr$ is never propagated past the operator that first produces it.
-%
-%\subsection{G\"{o}del's $T$ in $\lamA$}
-%In Section 2.1, we showed an encoding of natural numbers in $\lamA$, $\psit{nat}$.
-It would be instructive now for the reader to review the examples in that section and derive the various components of the compilation judgement for them.
-For the two negative examples, it is interesting to see why they fail to compile -- the first does not pass kind checking while the second is a user-defined error caught by the definition of $\tvar{s}$.
-%
-%We can now complete the encoding of G\"{o}del's $T$ in $\lamA$. The missing piece is the natural recursor. For review, the typing rule for the natural recursor can be written (omitting contexts):
-%\begin{mathpar}
-%\inferrule{
-%	{e_{1}:\tvar{nat}}\\
-%	{e_{2}:\tau}\\
-%	{e_{3}:\tvar{nat}\rightarrow\tau\rightarrow{\tau}}
-%}{
-%	{\tt natrec}~e_{1}~e_{2}~e_{3} : \tau
-%}
-%\end{mathpar}
-%The corresponding operator definition in $\lamA$, written in a concrete syntax that omits keywords and ``{\sf else err}'' branches is:
-%\begin{verbatim}
-%def natrec := forall k. op(fun (d0 d1 d2 : Den) := 
-%  didxcase d0 of NAT => 
-%  if d0 = nat => 
-%  didxcase d1 of k => 
-%  didxcase d2 of Type[NAT]*(Type[k]*Type[k]) =>
-%  def t1 := typeof d1 in 
-%  def t2 := typeof d2 in 
-%  if fst idx t2 = nat =>
-%  def s := snd idx t2 in 
-%  if fst s = snd s => 
-%  if fst s = t1 => 
-%  [[(fix f (n :: int) :: rep t1 :=
-%       if n = 0 then valof(d1) 
-%       else valof(d2) (n - 1) (f (n - 1))
-%     ) valof(d0), t1]]) in x
-%\end{verbatim}
-%Here, we see that natrec is parameterized by the kind of the index of the type $\tau$ above, corresponding to \texttt{t1} with index kind \texttt{k}.
-The body of the definition amounts to checking that the equational relations between the types for the three arguments that are implied by the inductive definition actually hold.
-The implementation of the natural recursor is a standard recursive definition that mirrors the operational semantics usually given to it.
-If we call this specification $\psit{natrec}$, then the composition of the two specifications produces a specification that is sound and complete against $T$:
-%$$\psi_{T} := \subst{\psit{natrec}}{\tvar{x}}{\psit{nat}}$$
-%
-%One might ask at this point what the benefit is of using this more complex specification style?
-For simple examples like this, it is clearly more complex than necessary.
-However, we will see in Section 4 examples where a purely inductive specification of a language construct requires significantly more machinery than a functional specification.
-In cases where the simpler inductive style is convenient, on the other hand, a sound translation from an inductive specification language to a naive type-level specification would not be difficult to develop.
-Moreover, automatically extracting an efficient implementation from an inductive specification of a semantics is generally far more difficult.
-With type-level specification, we err on the side of greater control whenever safe.
-%\subsection{Safety of $\lamA$}
-%As mentioned in the introduction, evaluating complex type-level functions to verify and translate programs is fraught with theoretical peril.
-Both the decidability of type-checking and the metatheoretic and global safety properties of the language itself are at risk of disruption by unwise user specifications.
-Fortunately, the formulation of $\lamA$ allows us to prove two important safety theorems.
-%
-%The first guarantees that the compilation judgement of Figure 2, which includes the type checking and translation phases, is decidable.
-%\begin{theorem}[Decidability of Compilation]
-%There is a terminating decision procedure that can derive the compilation judgement for any term $\psi$ if it exists, or halt if it does not exist.
-%\end{theorem}
-%\noindent
-%The second guarantees that successful compilation results in a well-typed internal language term that is representationally consistent with its type assignment.
-%\begin{theorem}[Type-Preserving Compilation]
-%If 
-%$$\compiles{\psi}{\mu}{\psiidx}{\kappa}{\psirep}$$ 
-%then $\mtentails{\cdot}{\mtof{\mu}{\psirep}}$.
-%\end{theorem}
-%
-%\subsubsection{Metatheoretic Properties of the Type-Level Language}
-%There are three main lemmas that we must show to ensure that compilation is decidable.
-%\begin{lemma}[Decidability of Kind Checking]
-%Derivation of the kind assignment judgement $\tentailsXt{\psi}{\kappa}$ is decidable.
-%\end{lemma}
-%\begin{lemma}[Decidability of Internal Type Checking]
-%Derivation of the internal type checking judgement $\mtentailsXt{\mu}{\psi}$ is decidable if $\psi$ is a canonical form for kind $\kIType$.
-%\end{lemma}
-%\begin{lemma}[Strong Normalization to Canonical Form] If a closed type-level term $\psi$ is well-kinded, that is 
-% $\tentails{\cdot}{\cdot}{\cdot}{\tof{\psi}{\kappa}}$,  then $\tevalsX{\psi}{\psi'}{\Sigma}$ for some $\Sigma$ and $\psi'$ is a canonical form of kind $\kappa$.
-%\end{lemma}
-%\noindent
-%We omit statements of the canonical forms lemma and other minor lemmas due to space considerations.
-Lemma 3 is the most difficult to prove but also absolutely critical -- a type-level language that does not guarantee termination will necessarily lead to undecidable type checking.
-Our approach, which we do not have space to detail here, is based on the standard logical relations formulation for strong normalization in System F, with special considerations of type indices.
-The check for strictly positive occurrences of the kind $\kTypeBlur$ ensures that values of kind $\kTypeBlur$ cannot sneak self-references into functions by clever use of {\sf idxcase}, by an argument exactly analagous to those given for total functional programming languages with inductive data types.
-In fact, we'll see in the next section that this analogy between type constructors and inductive data types has significance in practice.
-%
-%Using these three lemmas, we can show by simple inductive argument that the verification and compilation rules of Figure 5 are decidable, and thus that Theorem 1 holds.
-%\subsubsection{Representational Consistency}
-%Theorem 2 can be shown straightforwardly by noting that the big-step evaluation rules for denotations check representational consistency directly during normalization steps.
-The error context $\Sigma$ tracks whether any denotation constructed during the normalization and verification and translation processes is representationally inconsistent.
-If so, the final evaluation judgement cannot be derived.
-Combined with the representational consistency of the rule for arrow types, described in Section 2.2.2, we have a stronger theorem than the one stated -- that no denotation created during the compilation of the program $\psi$ is representationally inconsistent.
-%
-%The issue is that this form of checking amounts to ``dynamic checking'' of this particular invariant at the type-level.
-In particular, we cannot check that a well-kinded specification will produce a valid denotation or $\terr$ for all possible programs that it is applied to -- only those branches of the specification that are exercised for a particular program are checked.
-This means that users will occasionally be faced with specification errors even after the kind checking phase, in addition to any errors arising from problems in their programs.
-%
-%The reason we can't statically check specifications for this error is that it is a relational property between the two components of a denotation.
-As such, it requires much more sophisticated support for equational reasoning in the kind system.
-In $\psit{nat}$, for example, proving that even $\tvar{z}$ uses a representationally consistent denotation requires static knowledge that $\trepof{\tvar{nat}}=\tILtint$.
-While this is easily discernable from the specification as given, the simply-kinded system given does not track enough information to prove this kind of property.
-The problem is more clearly evident in the specifications for $\tvar{s}$ and $\tvar{natrec}$, which contain uses of \textsf{valof}.
-To prove that the internal term $1 + \mvalof{\psi}$ is well-typed without evaluating $\psi$ requires a proof that $\trepof{\ttypeof{\psi}}=\tILtint$.
-To prove that the denotation is representationally consistent then requires the same proof that $\trepof{\tvar{nat}}=\tILtint$.
-In Section 3, we give a method for translating type-level terms into Coq, which supports such reasoning.
-In practice, coverage tests to ensure that all branches of a specification are exercised by tests may be sufficient for prototype implementations of language features, with rigorous equational reasoning reserved for those features being deployed more widely or in safety-critical applications.
-Indeed, even in certified compilers like CompCert, the type checking and early translation phases are written in a general-purpose functional language (e.g.
-Ocaml) for practicality \cite{Leroy-Compcert-CACM}.
-%
-%\subsubsection{Local vs. Global Safety Guarantees}
-%While Theorem 2 guarantees that any global properties guaranteed by the internal type system will hold for all compiled programs, this mechanism does not support the introduction of new global safety properties for all possible programs.
-For instance, although the implementation of System $T$ described here can be shown sound and complete, some of the safety properties guaranteed by the constructs of $\psi_{T}$ hold only locally -- in program fragments that only use operators defined in $\psi_{T}$.
-A user could easily introduce a new constant $\tvar{b} := \tconst{\tden{-1}{\tvar{nat}}}$, for example.
-The introduction of this constant invalidates the mapping between the natural numbers of mathematics and the type $\tvar{nat}$, effectively redefining it.
-%
-%The purpose of this section was to demonstrate a technique that allows for the concrete specification and modular introduction of new primitives into a language.
-This facilitates experimentation with collections of primitives known (or at least strongly suspected) to be mutually consistent.
-But just as introducing new axioms into a proof assistant can cause problems if not done carefully, composing primitives should also be done with care.
-The hope is that by making experimentation as easy as importing a library and ensuring that it cannot invalidate fundamental properties of the language, different communities will more quickly converge upon powerful collections of mutually consistent primitives.
-%
-%In the next section, we sketch a general technique for proving stronger safety and composability theorems about specifications.
-%\section{Strengthening Specifications with Dependent Kinds}
-%In the previous section, we gave a minimal calculus that expresses the main ideas of this paper -- a type-level specification language that contains a uniform representation for types and operators.
-Operators were implemented as type-level functions operating on both indexed types and denotations.
-Although the language and compilation process could be shown to be safe in the face of these extensions, the extensions themselves could have errors in them -- we discussed representational consistency, soundness with respect to an independent specification (e.g.
-mathematical natural numbers) and composability of global guarantees.
-%
-%These issues all have in common that they require making complex statements about specifications.
-Proof assistants based in type theory such as Coq, Agda, Twelf and others are often used by programming languages researchers for just that task.
-By taking advantage of dependent types married with flexible inductive datatypes, these proof assistants are emerging as powerful tools for certified programming.
-They also possess the fundamental properties required of our metalanguage -- strong normalization.
-%
-%To practically realize a language featuring type-level specifications and demonstrate techniques that could be used for certifying these specifications, we propose lifting one of these languages into the type level.
-The key issue is that these languages do not admit a simple embedding for uniform type constructors.
-In most language specifications, types are represented using an inductive data type with a finite set of cases.
-In $\lamA$ however, types can be introduced anywhere.
-This is an instantiation of the well-known expression problem of Wadler \cite{wadler1998expression}.
-%
-%A number of techniques have been proposed in the literature to resolve this issue in general-purpose functional languages.
-Open data types are one popular approach \cite{conf/ppdp/LohH06}.
-The indexed types in $\lamA$ can be considered an open data kind where each possible index kind generates a new constructor.
-Each constructor must contain a slot to place the associated internal type's representation.
-We use this observation to construct a translation from a language in the style of $\lamA$ to Gallina, the dependently-typed language of the Coq proof assistant.
-The kinds that are possible for indices can be read statically from a type-level term.
-Index case statements containing closed kinds can thus be translated into \texttt{match} statements containing two cases: one for to the corresponding constructor, and one matching all other constructors.
-Index case statements containing kind variables (e.g.
-natrec) require a more sophisticated approach, however.
-The case kind must be unified against all possible closed index kinds and a match case must be generated for each one where unification succeeds.
-%
-%All the remaining constructs needed by $\lamA$ can be expressed using standard inductive data types.
-By also implementing the language and compiler itself in Gallina and providing a specification of the semantics of the internal language, type-level terms can be used to reason about the behavior of operator implementations in nearly arbitrary detail.
-A proof-passing approach to representational consistency as discussed in Section 2.4.3  can guarantee that users will only ever see errors originating in their own expressions because specifications themselves have been modularly verified.
-%
-%To prove strong composability results about a collection of types and primitives, there are a number of possible approaches.
-For the primitives of $\psi_{T}$, for example, it can be seen that it does not interfere with other primitives -- all denotations are assigned either a newly introduced type -- $\tvar{nat}$ -- or an external type -- ${\tt t1}$ -- paired with an internal term taken directly from an input denotation to the operator and also having type {\tt t1}, which we can inductively assume is semantically consistent.
-Proving such a theorem, while not trivial, is well within the capabilities of Coq and other proof assistants.
-Our choice of Coq is largely motivated by familiarity as well as Coq's sophisticated program extraction capabilities.
-%
-%\section{Examples}
-%We now give several examples of more sophisticated language constructs and how they may be implemented using this approach.
-We assume a type-level specification language structured like $\lamA$ but enriched with general inductive data kinds, though not necessarily dependent kinds.
-We assume kind inference when it is known to be decidable in System F (e.g.
-index kind declarations are omitted.) We also assume standard structures common to typed intermediate languages are available in the internal language.
-As in Section 2, missing else clauses are assumed to return \texttt{err}.
-%
-%\subsection{Record and Object Types}
-%A record type can be characterized by a collection of static field names mapped to other types.
-They are commonly implemented using some flat, heterogenous, untagged data structure, because the offsets can be computed statically.
-We can therefore declare the type family of records as a type-level function that produces a type indexed by this specification and represented by a generic flat heterogeneous structure.
-%
-%\begin{verbatim}
-%fun (spec : list[(str, Type)]) := 
-%  type[RECORD spec] :: flat[map spec 
-%    fun (s:str, t:Type) => (rep t)]
-%\end{verbatim}
-%
-%To extract a value from a record, we need an appropriate operator.
-We can implement this as a type-level function that produces an operator when given an index string.
-We assume an isRec function has been defined to check that the representation type is consistent with the index as defined above.
-We also assume that type-level numbers can be translated to internal language numbers using the \verb|#| operator.
-Several other helper functions are also assumed where the meaning is apparent.
-%
-%\begin{verbatim}
-%fun get(idx : str) := op(fun d : Den.
-%  didxcase d of RECORD =>
-%  if isRec typeof d =>
-%  def spec = open idx typeof d
-%  if containsIdx spec idx =>  
-%    [[valof(d)[#(getOffset spec idx)], 
-%      typeForIdx spec idx]])
-%\end{verbatim}
-%
-%An analagous set function can be implemented.
-Object types are like records, except they have additional metadata expressing inheritance relationships and other annotations.
-This additional information could also be packaged into the type index in a straightforward way, allowing for languages supporting a variety of different object protocols.
-%
-%\subsection{Scientific Units}
-%An example of a more specialized application domain is scientific unit checking.
-Numerical values (and sometimes collections of numerical values) can be annotated with dimensions (e.g.
-mass) and specific units of measure (e.g.
-grams).
-Addition and subtraction must operate on values with identical dimension and respect conversion rules if operating on heterogeneous units (e.g.
-grams + kilograms).
-Multiplication and division, on the other hand, cause the formation and elimination of compound units (e.g.
-meters/sec.).
-Although standalone type systems have been developed which respect these rules (e.g.
-\cite{conf/cefp/Kennedy09}), they are quite difficult to implement in terms of conventional language features, particularly if users hope to erase these annotations at run-time for performance reasons.
-Few languages have enough scientific users to justify adding this feature to the core language however.
-%
-%We can represent compound scientific types as being indexed by type-level lists of multiplicand and divisor units, where units are type-level values associated with dimensions.
-The corresponding rules can then be encoded directly as operators.
-%
-%\begin{verbatim}
-%def unitNum = fun (mult div : list[SIUnit]) :=
-%  type[SINUM mult, div] :: int
-%\end{verbatim}
-%\noindent
-%We omit the details of the operator implementations.
-%
-%\subsection{Length-Indexed Lists}
-%Length-indexed lists are a common example for dependently-typed languages.
-With type-level numbers, such lists are relatively straightforward to encode.
-All operators use standard arithmetic operations on the index to ensure that the length is accurately maintained.
-%
-%\begin{verbatim}
-%def list = fun (length : nat, t : Type) := 
-%  type[LIST length, t] :: array[rep t]
-%\end{verbatim}
-%
-%Other common examples of dependently-typed data structures can be encoded similarly, though the dependency in this case is only on type-level values (or perhaps denotations), but not run-time values.
-%
-%\subsubsection{Interoperability Layers}
-%A type-safe interoperability layer can be derived from an unsafe interoperability layer encoded by implementing the type system of the target language, with type representations corresponding to the core data structures available in the unsafe interoperability layer.
+
+\subsubsection{Active Typechecking}
+During the typechecking phase, the type of the primary operand, as determined by this dispatch protocol, is responsible for assigning a type to the expression as a whole.
+Let us consider the \verb|map| function from Listing \ref{map} once again.
+When it is compiled on Line 9 of Listing \ref{mapadd5dbl}, its first two argument types are given as \verb|global_ptr(double)|.
+As described in Section \ref{types}, this type, abbreviated \verb|A|, is an instance of \verb|ace.OpenCL.GlobalPtrType| which inherits from \verb|ace.OpenCL.PtrType| and ultimately from \verb|ace.Type|.
+So when the compiler encounters the expression \verb|input[gid]| on Line 6, it follows the dispatch protocol just described and assigns responsibility over typechecking it to \verb|A|.
+This is done by calling the \verb|resolve_|$X$ method of the responsible type, where $X$ is the syntactic form of the expression.
+In this case, the expression is of the \verb|Subscript| form, so the compiler calls \verb|A.resolve_Subscript|.
+
+The relevant portion of \verb|ace.OpenCL.GlobalPtrType| is shown in Listing \ref{pointers}.
+The \verb|verify_Subscript| method on Line 8 receives a context and the syntax tree of the node itself as input.
+The context contains information about other variables in scope, as well as other potentially relevant information, and also contains a method, \verb|resolve_type|, that can be used recursively resolve the types of subexpressions.
+On Line 9, this method is used to resolve the type of the slice subexpression, \verb|gid|, which is the machine-dependent integer type \verb|size_t| as discussed in Section II.E.
+On Line 10, it confirms that this type is an instance of an integer type.
+Thus, it assigns the whole expression, \verb|input[gid]|, the target type of the pointer, \verb|double|.
+Had a user attempted to index \verb|input| using a non-integer value, the method would take the other branch of the conditional and raise a type error with a relevant user-defined error message on Line 13.
+
+\subsubsection{Active Translation}
+Once typechecking a method is complete, the compiler must subsequently translate each Ace source expression into an expression in the target language, OpenCL in the examples thus far.
+It does so by again applying the dispatch protocol and calling a method of the form \verb|translate_|$X$, where $X$ is the syntactic form of the expression.
+This method is responsible for returning a copy of the expression's ast node with an additional attribute, \verb|code|, containing the source code of the translation.
+In this case, it is simply a direct translation to the corresponding OpenCL attribute access (Line 20), using the recursively-determined translations of the operands (Lines 16-17).
+More sophisticated abstractions may insert arbitrarily complex statements and expressions during this phase.
+The context also provides some support for non-local effects, such as new top-level declarations (not shown.)
+
+\subsection{Active Backends}\label{backends}
+Thus far, we have discussed using OpenCL as a backend with Ace.
+The OpenCL extension is the most mature as of this writing.
+However, Ace supports the introduction of new backends in a manner similar to the introduction of new types, by extending the \verb|clq.Backend| base class.
+Backends are provided as the first argument to the \verb|@clq.fn| decorator, as can be seen in Figure \ref{map}.
+Backends are responsible for some aspects of the grammar that do not admit simple dispatch to the type of a subterm, such as number and string literals or basic statements like \verb|while|.
+
+In addition to the OpenCL backend, preliminary C99 and CUDA backends are available (with the caveat that they have not been as fully developed or tested as of this writing.) Backends not based on the C family are also possible, but we leave such developments for future work.
+
+\subsection{Use Cases}
+The development of the full OpenCL language using only the extension mechanisms described above provides evidence of the power of this approach.
+Nothing about the core language was designed specifically for OpenCL.
+However, to be truly useful, as described in Sections \ref{multiparadigm} and  \ref{extensions}, the language must be able to support a wide array of primitive abstractions.
+We briefly describe a number of other abstractions that may be possible using this mechanism.
+Many of these are currently available either via inconvenient libraries or in standalone languages.
+With the Ace extension mechanism, we hope to achieve robust, natural implementations of many of these mechanisms within the same language.
+
+\paragraph{Partitioned Global Address Spaces}
+A number of recent languages in high-performance computing have been centered around a partitioned global address space model, including UPC, Chapel, X10 and others.
+These languages provide first-class support for accessing data transparently across a massively parallel cluster, which is verbose and poorly supported by standard C.
+The extension mechanism of Ace allows inelegant library-based approaches such as the Global Arrays library to be hidden behind natural wrappers that can use compile-time information to optimize performance and verify correctness.
+We have developed a prototype of this approach using the C backend and hope to expand upon it in future work.
+
+\paragraph{Other Parallel Abstractions}
+A number of other parallel abstractions, some of which are listed in \ref{multiparadigm}, also suffer from inelegant C-based implementations that spurred the creation of standalone languages.
+A study comparing a language-based concurrency solution for Java with an equivalent, though less clean, library-based solution found that language support is preferable but leads to many of the issues we have described \cite{cave2010comparing}.
+The extension mechanism is designed to enable library-based solutions that operate as first-class language-based solutions, barring the need for particularly exotic syntactic extensions.
+
+\paragraph{Domain-Specific Type Systems}
+Ace is a statically-typed language, so a number of domain-specific abstractions that promise to improve verifiability using types, as discussed in Section \ref{verifiability}, can be implemented using the extension mechanism.
+We hope that this will allow advances from the functional programming community to make their way into the professional end-user community more quickly, particularly those focused on scientific domains (e.g.
+\cite{conf/cefp/Kennedy09}).
+
+\paragraph{Specialized Optimizations}
+In many cases, code optimization requires domain-specific knowledge or sophisticated, parametrizable heuristics.
+Existing compilers make implementing and distribution such optimizations difficult.
+With active libraries in Ace, optimizations can be distributed directly with the libraries that they work with.
+For instance, we have implemented substantial portions of the NVidia GPU-specific optimizations described in \cite{yang2010gpgpu} as a library that uses the extension mechanism to track affine transformations of the thread index used to access arrays, in order to construct a summary of the memory access patterns of the kernel, which can be used both for single-kernel optimization (as in \cite{yang2010gpgpu}) and for future research on cross-kernel fusion and other optimizations.
+
+\paragraph{Instrumentation}
+Several sophisticated feedback-directed optimizations and adaptive run-time protocols require instrumenting code in other ways.
+The extension mechanism enables granular instrumentation based on the form of an operation as well as its constituent types, easing the implementation of such tools.
+This ability could also be used to collect data useful for more rigorous usability and usage studies of languages and abstractions, and we plan on following up on this line of research going forward.
+
+%
+%\section{Case Study: Neurobiological Circuit Simulation}
+%An important criteria that practitioners use to evaluate a language or abstraction, as discussed in Section \ref{social}, is whether significant case studies have been conducted with it.
+In this section, we briefly (due to space limitations) discuss an application of the Ace OpenCL library, Python host bindings and code generation features for developing a modular, high-performance scientific simulation library used to simulate  thousands of parallel realizations of a spiking neurobiological circuit on a GPU.
+%
+%\subsection{Background}
+%A neural circuit can be modeled as a network of coupled differential equations, where each node corresponds to a single neuron.
+Each neuron is modeled using one or more ordinary differential equations.
+These equations capture the dynamics of physically important quantities like the cell's membrane potential or the conductance across various kinds of ion channels and can take many forms \cite{neurobook}.
+Single simulations can contain from hundreds to tens of millions of neurons each, depending on the specific problem being studied.
+In some cases, such as when studying the effects of noise on network dynamics or to sweep a parameter space, hundreds or thousands of realizations must be generated.
+In these cases, care must be taken to only probe the simulation for relevant data and process portions of it as the simulation progresses, because the amount of data generated is often too large to store in its entirety for later analysis.
+%
+%The research group we discuss here (of which the first author was a member) was studying a problem that required running up to 1,000 realizations of a network of between 4,000 and 10,000 neurons each.
+An initial solution to this problem used the Brian framework, written in Python, to conduct these simulations on a CPU.
+Brian was selected because it allowed the structure of the  simulation to be specified in modular and straightforward manner.
+This solution required between 60 and 70 minutes to conduct the simulations and up to 8 hours to analyze the data each time a parameter of the simulation was modified.
+%
+%Unsatisfied with the performance of this approach, the group developed an accelerated variant of the simulation using C++ and CUDA.
+Although this produced significant speedups, reducing the time for a simulation by a factor of 40 and the runtime of the slowest analyses by a factor of 200, the overall workflow was also significantly disrupted.
+In order to support the many variants of models, parameter sets, and probing protocols, C preprocessor flags were necessary to selectively include or exclude code snippets.
+This quickly led to an incomprehensible and difficult to maintain file structure.
+Moreover, much of the simpler data analysis and visualization was conducted using Python, so marshalling the relevant data between processes also became an issue.
+%
+%\subsection{The {\sf cl.egans} Simulation Library}
+%In order to eliminate these issues while retaining the performance profile of the GPU-accelerated code, the project was ported to Ace.
+Rather than using preprocessor directives to control the code contained in the final GPU kernels used to execute the simulation and data analyses, the group was able to develop a more modular  library called {\sf cl.egans}\footnote{...after {\it c.
+elegans}, a model organism in neuroscience} based on the language's compile-time code generation mechanism and Python and OpenCL bindings.
+%
+%{\sf cl.egans} leverages Python's object-oriented features to enable modular, hierarchical simulation specifications.
+For example, Figure \ref{spec} shows an example where a neuron model (\verb|ReducedLIF|) is added to the root of the simulation, a synapse model (\verb|ExponentialSynapse|) is then added to it, and its conductance is probed in the same way, by adding a probe model as a child of the synapse model.
+If interleaved analysis needed to be conducted as well, it would be specified in the same way.
+%
+%Implementations of these classes do not evaluate the simulation logic directly, but rather contain methods that generate Ace source code for insertion at various points, called {\it hooks}, in the final simulation kernel.
+The hook that code is inserted into is determined by the method name, and code can be inserted into any hook defined anywhere upstream in the simulation tree.
+New hooks can also be defined in these methods and these become available for use by child nodes.
+Figure \ref{impl} shows an example of a class that inserts code in the \verb|model_code| hook and defines several new hooks.
+This protocol is closely related to the notion of {\it frame-oriented programming}.
+Although highly modular, this strategy avoids the performance penalties associated with standard object-oriented methodologies via code generation.
+%
+%Compared to a similar protocol targeting OpenCL directly, the required code generation logic is significantly simpler because it enables classes like \verb|StateVariable| to be written generically for all types of state variables, without carrying extra parameters and {\it ad hoc} logic to extract and compute the result types of generated expressions.
+Moreover, because types are first-class objects in the metalanguage, they can be examined during the memory allocation step to enable features like fully-automatic parallelization of multiple realizations across one or more devices, a major feature of {\sf cl.egans} that competing frameworks cannot easily offer.
+%
+%
+%Once the kernel has been generated and memory has been allocated, the simulation can be executed directly from Python using the bindings described in Section \ref{direct}.
+The results of this simulation are immediately available to the Python code following the simulation and can be visualized and further analyzed using standard tools.
+Once the computations are complete, the Python garbage collector is able to handle deallocation of GPU memory automatically (a feature of the underlying \verb|pyopencl| library \cite{klockner2011pycuda}.)
+%
+%Using this Ace-based framework, the benefits of the Brian-based workflow were recovered without the  corresponding decrease in performance relative to the previous CUDA-based solution, leading ultimately to a satisfying solution for the group conducting this research.
+%
+%\begin{figure}
+%\lstinputlisting{listing10.py}
+%\caption{\texttt{[listing10.py]} An example of a nested simulation tree, showing that specifying a simulation is both simple and modular.
+The first argument to the constructor specifies each node's parent.}
+%\label{spec}
+%\end{figure}
+%
+%\begin{figure}
+%\lstinputlisting{listing11.py}
+%\caption{\texttt{[listing11.py]} An example of a hook that inserts code and also inserts new, nested hooks for downstream simulation nodes  below that.}
+%\label{impl}
+%\end{figure}
+%
 %\section{Related Work}
+%\subsection{Active Libraries in Ace}
+%Libraries that have such capabilities has been called {\it active libraries} in prior proposals \cite{activelibraries}.
+A number of  projects, such as Blitz++, have taken advantage of the C++ preprocessor and template-based metaprogramming system to implement domain-specific optimizations.
+In Ace, we replace these brittle mini-languages with a general-purpose language.
+This allows for several interesting uses that we discuss in the following sections.
+%
+%
+%\subsubsection{Structural Polymorphism}
+%In Section \ref{hof}, we discussed several strategies for achieving {\it polymorphism} -- the ability to create functions and data structures that operate over more than a single type.
+In Ace, all functions are implicitly polymorphic and can be called with arguments of {\it any type that supports the operations used by the function}.
+For example, in Figure \ref{map}, \verb|in| can be any type that supports indexing by a variable of type \verb|size_t| to produce a value of a type that can be passed into \verb|fn|, which must then produce a value consistent with indexing into \verb|out|.
+OpenCL pointer types are consistent with these constraints, for example.
+Although powerful, this also demonstrates a caveat of this approach -- that it is more difficult to give a function a concise signature, because arguments are constrained by capability, rather than to a single type \cite{malayeri2009structural}.
+%
+%Structural typing can be compared to the approach taken by dynamically-typed languages that rely on ``duck typing''.
+It is more flexible than the parametric polymorphism found in many functional languages and in languages like Java (which only allow polymorphic functions that are valid for {\it all} possible types), but is of comparable strength to the template system found in C++.
+It can be helpful to think of each function as being preceded by an implicit template header that assigns each argument its own unique type parameter.
+At function call sites, these parameters are implicitly specialized with the types of the provided arguments.
+This choice is again motivated by the criteria of conciseness given in Section \ref{syntax}.
+%
 %\subsection{Type-Level Computation} %Haskell, Ur and $\Omega$mega
-%System $\text{F}_{\omega}$ with simple case analysis provides the basis of type-level computation in Haskell (where type-level functions are called type families \cite{Chakravarty:2005:ATC}).
+%System XX with simple case analysis provides the basis of type-level computation in Haskell (where type-level functions are called type families \cite{Chakravarty:2005:ATC}).
 Ur uses type-level records and names to support typesafe metaprogramming, with applications to web programming \cite{conf/pldi/Chlipala10}.
@@
 %
-%\subsection{Language Frameworks}
-%When the mechanisms available in an existing language prove insufficient, researchers and domain experts must design a new language.
+\subsection{Language Frameworks}
+When the mechanisms available in an existing language prove insufficient, researchers and domain experts must design a new language.
 A number of tools have been developed to assist with this task, including compiler generators, language workbenches and domain-specific language frameworks (cf \cite{fowler2010domain}).
-%
-%A major barrier to adoption is the fact that interoperability is intrinsically problematic.
+
+A major barrier to adoption is the fact that interoperability is intrinsically problematic.
 Even languages which target a common platform, such as the Java Virtual Machine, can only interact using its limited set of primitives.
@@
 Java).
-%
-%Instead of focusing on defining standalone languages, type-level specification gives greater responsibility in a granular manner to libraries.
+
+Instead of focusing on defining standalone languages, type-level specification gives greater responsibility in a granular manner to libraries.
+In this way, a range of constructs can coexist within the same program and, assuming that it can be shown by some method that various constructs are safely composable, be mixed and matched.
+The main limitation is that the protocol requires defining a fixed source grammar, whereas a specialized language has considerable flexibility in that regard.
+Nevertheless, as Ace shows, a simple grammar can be used quite flexibly.
+\subsection{Extensible Compilers}
+An alternative methodology is to implement language features granularly as compiler extensions.
+As discussed in Section 1, existing designs suffer from the same problems related to composability, modularity\-, safety and security as extensible languages, while also adding the issue of language fragmentation.
+
+Type-level specification can in fact be implemented within a compiler, rather than provided as a core language feature.
+This would resolve some of the issues, as described in this paper.
+However, by leveraging type-level computation to integrate the protocol directly into the language, we benefit from common module systems and other shared infrastructure.
+We also avoid the fragmentation issue.
+\subsection{Specification Languages}
+Several {\it specification languages} (or {\it logical frameworks}) based on these theoretical formulations exist, including the OBJ family of languages (e.g.
+CafeOBJ \cite{Diaconescu-Futatsugi01}).
+They provide support for verifying a program against a language specification, and can automatically execute these programs as well in some cases.
+The  language itself specifies which verification and execution strategies are used.
+
+Type-determined compilation takes a more concrete approach to the problem, focusing on combining {\it implementations} of different\- logics, rather than simply their specifications.
+In other words, it focuses on combining {\it type checkers} and {\it implementation strategies} rather than more abstract representations of a language's type system and dynamic semantics.
+In Section 4, we outlined a preliminary approach based on proof assistant available for the type-level language to unify these approaches, and we hope to continue this line of research in future work.
+
+\section{Conclusion}
+In addition to the novel architecture of Ace as a whole, we note several individually novel features introduced here: %in this paper:
+
+\begin{itemize}
+\item The \ATT~mechanism, which is a generalization of the concept of active libraries \cite{activelibraries}  where types are metalanguage objects.
+\item The method Ace uses to eliminate the need for type annotations in most cases, which combines a form of type inference with type propagation.
+\item The method Ace uses check correctness of generated code by checking representational consistency constraints associated with types, detailed in Section \ref{repcon}. 
+\item The type-aware simulation orchestration techniques used in the \texttt{cl\_egans} library, described in Section \ref{clegans}.
+\end{itemize}
+
+Readers familiar with the Python programming language will recognize the style of syntax used in Figure \ref{map}.
+In fact, Ace uses the Python grammar and parsing facilities directly.
+Several factors motivated this design decision.
+First, Python's syntax is widely credited as being particularly simple and readable, due to its use of significant whitespace and conventional mathematical notation.
+Python is one of the most widely-used languages in scientific computing, so its syntax is already familiar to much of the field.
+And significantly, a large ecosystem of tools already exist that work with Python files, such as code editors, syntax highlighters, style checkers and documentation generators.
+These can be used without modification to work with Ace files.
+Therefore, by re-using an existing, widely-used grammar, we are able to satisfy many of the design criteria described in Section \ref{syntax} and the adoption criteria described in Section \ref{tools} without significant development effort.
+
+Professional end-users demand much from new languages and abstractions.
+In this paper, we began by generating a concrete, detailed set of design and adoption criteria that we hope will be of broad interest and utility to the research community.
+Based on these constraints, we designed a new language, Ace, making several pragmatic design decisions and utilizing advanced techniques, including type inference, structural typing, compile-time metaprogramming and active libraries, to uniquely satisfy many of the criteria we discuss, particularly those related to extensibility.
+We validated the extension mechanism with a mature implementation of  the entirety of the OpenCL type system, as well as preliminary implementations of some other features.
+Finally, we demonstrated that this language was useful in practice, drastically improving performance without negatively impacting the high-level scientific workflow of a large-scale neurobiological circuit simulation project.
+Going forward, we hope that Ace (or simply the key techniques it proposes, by some other vehicle) will be developed further by the community to strengthen the foundations upon which new abstractions are implemented and deployed into professional end-user development communities.
+
+\section{Availability}
+Ace is available under the LGPL license and is developed openly and collaboratively using the popular Github platform at \url{https://github.com/cyrus-/ace}.
+Documentation, examples and other learning materials will be available at \url{http://acelang.org/}.
+{\color{red}(by the time of the conference)}
+
+\section{Acknowledgments}
+CO was funded by the DOE Computational Science Graduate Fellowship under grant number DE-FG02-97ER25308. NF was funded by ...
+
+\section{Background}
+The approach we describe, {\it active type-checking and translation} (AT\&T), makes use of type-level computation in a novel way.
+To review, in languages supporting type-level computation, the syntactic class of types is not simply declarative.
+Instead, it forms a programming language itself (the {\it type-level language}).
+Types themselves are one kind of value in this language, but there can be many others.
+To ensure the safety of type-level computations, {\it kinds} classify type-level terms, just as types classify expression-level terms.
+The simplest example of a language featuring type-level computation is Girard's System $\text{F}_{\omega}$ \cite{tapl}.
+In $\text{F}_{\omega}$, types have kind $\star$ and type-level functions have kinds like $\star\rightarrow\star$.
+A growing number of implemented languages now feature more sophisticated type-level languages (see Section 5).
+We emphasize that type-level computation occurs during compilation, rather than at run-time, because type-level terms that are used where types would normally be expected must be reduced to normal form before type-checking can proceed.
+
+In this work, we wish to allow extensions to strengthen the static semantics of our language.
+Naturally, extension specifications will also need to be evaluated during compilation and manipulate representations of types.
+This observation suggests that the type-level language may be able to serve directly as a specification language.
+In this paper, we show that this is indeed the case.
+By introducing some new constructs at the type-level, developers can specify the semantics of operators associated with newly-introduced families of primitive types with type-level functions.
+The compiler front-end invokes these functions to synthesize types for and assign meanings to expressions, by translation into a {\it typed internal language}.
+Unlike conventional metaprogramming systems, these {\it type-level specifications} do not directly manipulate or rewrite expressions.
+Instead, they examine and manipulate the types of these expressions.
+By using a sufficiently constraining kind system and incorporating techniques from typed compilation into the type-level language directly, the global safety properties of the language and compilation process can be guaranteed.
+In other words, users can only {\it increase} the safety of the language.
+
+We focus on extending the static semantics of a language with a fixed, though flexible, grammar.
+Techniques for extensible parsing have been proposed in the past (e.g.
+\cite{journals/entcs/BrabrandSV03}), and we conjecture that these can be made compatible with the mechanism described in this paper with some simple modifications, but we do not discuss this further here.
+We also focus on {\it functional}, rather than declarative, specifications of language constructs.
+Extracting a compiler from a declarative language specification (e.g.
+in Twelf \cite{twelf}) has not yet been shown practical, but we note that a future mechanism of this sort could safely target a language implementing the mechanism we discuss here.
+
+\section{Type-Level Specifications in $\lamA$}
+\subsection{Example: Natural Numbers in $\lamA$} % G\"{o}del's $T$ 
+We begin with a simple calculus with no primitive notion of natural numbers, nor any more general notion of an inductive data type.
+We can, however, concretely specify both the static and dynamic semantics of natural numbers, including the natural recursor of G\"{o}del's $T$ \cite{tapl}, using type-level specifications.
+Let us begin in Figure 1 with the type $\tvar{nat}$ and its constructors, $\tvar{z}$ and $\tvar{s}$.
+%\begin{figure}
+%$\tfam{NAT}{\kunit}{
+%	\tlam{\tof{\tvar{self}}{\kType{\fvar{NAT}}}}{\titype{\dint}}}{(\\
+%	~~~~
+%	\topp{\tops{z}{\tlam{\tof{\tvar{self}}{\kTypeBlur}}{\tconst{\tden{\titerm{0}}{\tvar{self}}}}}}{
+%	\\~~~~
+%	\tops{s}{\tlam{\tof{\tvar{self}}{\kTypeBlur}}{\tOp{
+%		\tlam{\tof{\tvar{d1}}{\kDen}}{
+%		\\~~~~~~~~~\tifeq{\ttypeof{\tvar{d1}}}{\tvar{self}}{\tden{\tvalof{\tvar{d1}}}{\tvar{self}}}{\terr}}}}}}
+%	)}{\Theta}{{\textsf{let~}}{\text{nat}=\ttype{\tunit}{\fvar{nat}}}{\textsf{~in~}}{x}}$
+%\caption{Specifying natural numbers in $\lamA$. This term is referred to as $\psit{nat}$.}
+%\end{figure}
+%\begin{figure*}
+%\small
+%$$\begin{array}{rccl}
+%\textbf{expressions} 				&	e	&	::=	&	\evar{x} \pipe 
+%														\elam{\evar{x}}{\psi}{e} \pipe 
+%														\eapp{e_{1}}{e_{2}} \pipe
+%												\eop{\psitype}{op}{
+%															\splat{\psi}{1}{m}
+%														}{
+%  												    		
+%														} \pipe
+%														\eop{\psitype}{op}{
+%															\splat{\psi}{1}{m}
+%														}{
+%  												    		\splat{e}{1}{n}
+%														} \\
+%									& 		&		& 	\\
+%							
+%\textbf{type-level specifications} 	& \psi 	& ::= 	& 	\tvar{t} \pipe 
+%														\tlam{\splatTwo{\tvar{t}}{{:}\kappa}{1}{n}}{\psi} \pipe 
+%														\tapp{\psi}{\splat{\psi}{1}{n}} \pipe
+%														\tifeq{\psi_{0}}{\psi_{1}}{\psi_{2}}{\psi_{3}} \\
+%												
+%\text{standard terms}	 			& 		& \pipe	& 	\tunit \pipe 
+%														\tpair{\psi_{1}}{\psi_{2}} \pipe 
+%														\tfst{\psi} \pipe 
+%														\tsnd{\psi} 
+%														\\
+%														
+%\text{type families}				&		& \pipe	& 	\tfam{fam}{\kappaidx}{\psirep}{\theta}{\Theta}{\psi}\pipe
+%														\tfamcase{\psi}{\phi}{\psi_0}{\psi_1}\\
+%
+%\text{types} 						& 		& \pipe	& 	\ttypestd \pipe 
+%														\tidx{\psitype} \pipe
+%														\trepof{\psitype}
+%														\\
+%																								
+%\text{operator definitions} 		& 		& \pipe & 	\tconst{\psi} \pipe 
+%														\tOp{\psi} \\
+%												
+%\text{denotations} 				& 		 & 	\pipe	&	\tden{\psiiterm}{\psitype} \pipe 
+%														\tvalof{\psiden} \pipe
+%														\ttypeof{\psiden} \pipe
+%														\terr \\
+%
+%\text{internal language}		&		&	\pipe	&	\titerm{\mu} \pipe \titype{\delta}\\
+%
+%\text{programs}					&		&	\pipe	&	\tprog{e}\\
+%
+%\text{family specifications}	&	\phi		&	::= & \fvar{fam} \pipe \tfamSpecStd\\
+%\text{operator lists}		&	\theta	&	::= &	\topsempty \pipe 
+%												\topp{\theta}{\tops{op}{\psi}}\\
+%												
+%\textbf{kinds} 					& \kappa	&	::=	&	\karrow{\splat{\kappa}{1}{n}}{\kappa} \pipe 
+%												\kunit \pipe 
+%												\kpair{\kappa_{1}}{\kappa_{2}} \pipe 
+%												\kType{\eta} \pipe 
+%												\kTypeBlur \pipe
+%												\kOp{n} \pipe
+%												\kDenk{\eta} \pipe 
+%												\kDen \pipe 
+%												\kIType \pipe \kITerm \pipe
+%												\kProg 
+%												\\
+%												
+%\text{family signatures}				& \eta		& 	::=	& \kFamStd\\
+%\text{operator list signatures}			& \Theta	&	::=	&	\cdot \pipe \Theta; \tvar{op}: \kappa\\
+%											 							&		&		&	\\
+%\textbf{internal terms} 				& 	\mu	&	::=	&	\ivar{u} \pipe 
+%												\ilam{\ivar{u}}{\delta}{\mu} \pipe 
+%												\ifix{\ivar{f}}{\delta}{\mu} \pipe 
+%												\iapp{\mu_{1}}{\mu_{2}} \pipe 
+%												\iIfEq{\mu_{0}}{\mu_{1}}{\mu_{2}}{\mu_{3}}\\
+%							& 		& 	\pipe	&	\iintlit \pipe 
+%												\iop{\mu_{1}}{\mu_{2}} \pipe 
+%												\ipair{\mu_{1}}{\mu_{2}} \pipe 
+%												\ifst{\mu} \pipe
+%												\isnd{\mu} \pipe
+%												\iup{\psi} 
+%												\\
+%\textbf{internal types}			&	\delta	&	::=	&    \darrow{\delta_1}{\delta_2} \pipe
+%												\dint \pipe
+%												\dpair{\delta_1}{\delta_2} \pipe
+%												\dup{\psi}\\
+%%\text{cases}					&\tcasev 	&	::=	&	\caseZ{\splat{\kappa}{1}{n}}{\psi} \pipe 
+%%												\caseS{\tcasev}{\tcasev}\\
+%%							&		&		&	\\
+%\end{array}$$
+%\caption{Syntax of $\lamA$. $\mathbb{Z}$ denotes integer literals and $\klabel{label}$ denotes any label.}
+%\end{figure*}
+%\begin{figure*}
+%\begin{mathpar}
+%\inferrule{
+%	\overbrace{\tentailst{\cdot}{\psiprog}{\kProg}}^{\text{\small kind checking}}{ }\\
+%	\overbrace{\tevalsX{\psiprog}{\tprog{e}}{\noprob}}^{\text{\small specification normalization}}{ }\\
+%	\overbrace{\translates{\cdot}{e}{\mu}{\ttypestd}}^{\text{\small verification \& translation}}{ }
+%}{
+%%	\compiles{\psiprog}{\mu}{\psiidx}{\kappa}{\psirep}
+%	\compiles{\psiprog}{\mu}{\ttypestd}
+%}
+%\end{mathpar}
+%\caption{Central compilation judgement of $\lamA$.}
+%\end{figure*}
+%%\noindent
+%%
+%%The \textsf{def} construct is the type-level analog of the standard \textsf{let} construct.
+Indeed, the entire term consists of type-level constructs.
+The definition of $\tvar{nat}$ says that it is a type indexed by a uniquely labeled unit (that is, it is a simple type) that will be represented in the internal language by integers.
+The constructor $\tvar{z}$ is declared as a constant, denoted by the value $0$ and assigned the type $\tvar{nat}$.
+The successor operator $\tvar{s}$ first checks its argument to ensure that it is a $\tvar{nat}$ (which requires first checking the index's kind to ensure that it has the correct label using the $\sf{idxcase}_{D}$ statement).
+If so, it returns a denotation that assigns the type $\tvar{nat}$ to an internal term that increments the value of the argument.
+Otherwise, the special denotation $\terr$ is returned.
+We will return to explain these constructs in more detail shortly.
+%%
+%%This term, which we will call $\psit{nat}$, is an open term because it has not been used to write a program yet.
+To use this specification to write a program, we substitute a \textsf{program} term containing an expression referring to these definitions for $\tvar{x}$.
+For example, the following program corresponds to the  number $1$:
+%$$\psi_{1} := \subst{\tprog{
+%	\eop{\text{nat}}{s}{}{\eop{\text{nat}}{z}{}{}}
+%}}{\psi_{\text{nat}}}$$
+%%\noindent
+%%This program will be compiled to the internal term $1 + 0$ and assigned the type $\tvar{nat}$.
+This notion is formalized by the {\it central compilation judgement} of $\lamA$, given in Figure 3.
+In this case, we will be able to derive that
+%$$
+%\compiless{\psi_{1}}{1 + 0}{\tvar{nat}}
+%$$
+%%Compilation will not succeed for invalid terms, e.g.
+%%\begin{eqnarray*}
+%%\psi_{-} & := & \subst{\tprog{\econst{\tvar{s}}}}{\tvar{x}}{\psi_{\text{nat}}}\\
+%%\psi_{-}' & := & \subst{\tprog{\eop{\tvar{s}}{\elam{\evar{n}}{\tvar{nat}}{\eop{\tvar{s}}{\evar{n}}}}}}{\tvar{x}}{\psi_{\text{nat}}}
+%%\end{eqnarray*}
+%%Indeed, compilation with $\psi_{\text{nat}}$ as given above admits all well-typed expressions of the simply-typed lambda calculus with natural numbers and no others, compiling $\tvar{nat}$s to internal terms that evaluate to their integer representations.
+%%
+%%\subsection{$\lamA$}
+%\begin{figure*}
+%$\fbox{\inferrule{}{\tentailsXt{\psi}{\kappa}}}$~
+%~~~~$\fCtx ::= \cdot \pipe \fCtx, \fSpecStd$
+%~~~~$\tvarCtx ::= \cdot \pipe \tvarCtx, \tof{\tvar{t}}{\kappa}$
+%\begin{mathpar}
+%\small
+%\inferrule{
+%	\tof{\tvar{t}}{\kappa} \in \tvarCtx
+%}{
+%  \tentailsXt{\tvar{t}}{\kappa}
+%}~(\text{var}_\psi)
+%
+%\inferrule{ }{
+%	\tentailsXt{\tunit}{\kunit}
+%}~(\text{unit}_\psi)
+%
+%\inferrule{
+%	\tentailsXt{\psi_{1}}{\kappa_{1}}\\
+%	\tentailsXt{\psi_{2}}{\kappa_{2}}
+%}{
+%	\tentailsXt{\tpair{\psi_{1}}{\psi_{2}}}{\kpair{\kappa_{1}}{\kappa_{2}}}
+%}~(\times_\psi)
+%
+%\inferrule{
+%	\tentailsXt{\psi}{\kpair{\kappa_{1}}{\kappa_{2}}}
+%}{
+%	\tentailsXt{\tfst{\psi}}{\kappa_{1}}
+%}~(\text{fst}_\psi)
+%
+%\inferrule{
+%	\tentailsXt{\psi}{\kpair{\kappa_{1}}{\kappa_{2}}}
+%}{
+%	\tentailsXt{\tsnd{\psi}}{\kappa_{2}}
+%}~(\text{snd}_\psi)
+%
+%\inferrule{
+%  \kentailsX{\kappa_1}\\
+%  \cdots\\
+%  \kentailsX{\kappa_n}\\
+%  \tentailst{\tCtxXT{\splatTwo{\tvar{t}}{{:}\kappa}{1}{n}}}{\psi}{\kappa}
+%}{
+%  \tentailsXt{\tlam{\splatTwo{\tvar{t}}{{:}\kappa}{1}{n}}{\psi}}{\karrow{\splat{\kappa}{1}{n}}{\kappa}}
+%}~\text{($\lambda_\psi$)}
+%
+%\inferrule{
+%  \tentailsXt{\psi}{\karrow{\splat{\kappa}{1}{n}}{\kappa}}\\
+%  \tentailsXt{\psi_{1}}{\kappa_{1}}\\
+%  \cdots\\
+%  \tentailsXt{\psi_{n}}{\kappa_{n}}
+%}{
+%  \tentailsXt{\tapp{\psi}{\splat{\psi}{1}{n}}}{\kappa}
+%}~(\text{app}_\psi)
+%
+%\inferrule{
+%	\tentailsXt{\psi_{0}}{\kappa_{1}}\\
+%	\tentailsXt{\psi_{1}}{\kappa_{1}}\\
+%	\tentailsXt{\psi_{2}}{\kappa_{2}}\\
+%	\tentailsXt{\psi_{3}}{\kappa_{2}}
+%}{
+%	\tentailsXt{\tifeq{\psi_{0}}{\psi_{1}}{\psi_{2}}{\psi_{3}}}{\kappa_{2}}
+%}~(\text{cond}_\psi)
+%
+%\inferrule{
+%	\tentailst{\tCtxXF{\fSpecStd}}{\tfamSpecStd}{\kFamStd}\\
+%	\tentailst{\tCtxXF{\tof{\fvar{fam}}{\kFamStd}}}{\psi}{\kappa}
+%}{
+%	\tentailsXt{\tfamStd}{\kappa}
+%}~(\text{fambind})
+%
+%\inferrule{
+%	\tentailsXt{\psitype}{\kTypeBlur}\\
+%	\tentailsXt{\phi}{\eta}\\
+%	\tentailsXt{\psi_0}{\karrow{\kType{\eta}}{\kappa}}\\
+%	\tentailsXt{\psi_1}{\kappa}
+%}{
+%	\tentailsXt{\tfamcase{\psitype}{\phi}{\psi_0}{\psi_1}}{\kappa}
+%}~(\text{famcase-T})
+%
+%\inferrule{
+%	\tentailsXt{\psiden}{\kDen}\\
+%	\tentailsXt{\phi}{\eta}\\
+%	\tentailsXt{\psi_0}{\karrow{\kDenk{\eta}}{\kappa}}\\
+%	\tentailsXt{\psi_1}{\kappa}
+%}{
+%	\tentailsXt{\tfamcase{\psiden}{\phi}{\psi_0}{\psi_1}}{\kappa}
+%}~(\text{famcase-D})
+%
+%\inferrule{
+%\tentailsXt{\phi}{\kFamStd}\\
+%\tentailsXt{\psiidx}{\kappaidx}
+%}{
+%	\tentailsXt{\ttype{\psiidx}{\phi}}{\kType{\kFamStd}}
+%} ~(\text{Type}_I)
+%
+%\inferrule{
+%	\tentailsXt{\psi}{\kType{\eta}}
+%}{
+%	\tentailsXt{\psi}{\kTypeBlur}
+%}~(\text{Type-$\subseteq$})
+%
+%\inferrule{
+%	\tentailsXt{\psi}{\kType{\kFamStd}}
+%}{
+%	\tentailsXt{\tidx{\psi}}{\kappaidx}
+%}~(\text{idxof})
+%
+%\inferrule{
+%	\tentailsXt{\psi}{\kTypeBlur}
+%}{
+%	\tentailsXt{\trepof{\psi}}{\kIType}
+%}~(\text{repof})
+%
+%\inferrule{
+%	\tentailsXt{\psi}{\kDen}
+%}{
+%	\tentailsXt{\tconst{\psi}}{\kOp{0}}
+%}~(\text{const})
+%
+%\inferrule{
+%	\tentailsXt{\psi}{\karrow{\kDen, \stackrel{n}{\cdots}, \kDen}{\kDen}}
+%}{
+%	\tentailsXt{\tOp{\psi}}{\kOp{n}}
+%}~(\text{$n$-op})
+%
+%\inferrule{
+%	\tentailsXt{\psiiterm}{\kITerm}\\
+%	\tentailsXt{\psitype}{\kType{\eta}}
+%}{
+%	\tentailsXt{\tden{\psiiterm}{\psitype}}{\kDenk{\eta}}
+%}~(\text{Den}_I)
+%
+%\inferrule{
+%	\tentailsXt{\psiiterm}{\kITerm}\\
+%	\tentailsXt{\psitype}{\kTypeBlur}
+%}{
+%	\tentailsXt{\tden{\psiiterm}{\psitype}}{\kDen}
+%}~(\text{Den}_I\text{-$\subseteq$})
+%
+%\inferrule{
+%	\tentailsXt{\psi}{\kDenk{\eta}}
+%}{
+%	\tentailsXt{\psi}{\kDen}
+%}~(\text{Den-$\subseteq$})
+%
+%\inferrule{ }{
+%	\tentailsXt{\terr}{\kDen}
+%}~(\text{err})
+%
+%\inferrule{
+%	\tentailsXt{\psi}{\kDenk{\eta}}
+%}{
+%	\tentailsXt{\ttypeof{\psi}}{\kType{\eta}}
+%}~(\text{typeof})
+%
+%\inferrule{
+%	\tentailsXt{\psi}{\kDenk{\eta}}
+%} {
+%	\tentailsXt{\tvalof{\psi}}{\kITerm}
+%}~(\text{valof})
+%\\
+%\inferrule{
+%	\ientails{\iMkCtx{\fCtx}{\tvarCtx}{\cdot}}{\mu}
+%}{
+%	\tentailsXt{\titerm{\mu}}{\kITerm}	
+%}~(\text{iterm})
+%
+%\inferrule{
+%	\tentailsX{\delta}
+%}{
+%	\tentailsXt{\titype{\delta}}{\kIType}
+%}~(\text{itype})
+%
+%\inferrule{
+%	\ientails{\iMkCtx{\fCtx}{\tvarCtx}{\cdot}}{e}
+%}{
+%	\tentailsXt{\tprog{e}}{\kProg}
+%}~(\text{program})
+%\end{mathpar}
+%$\fbox{\inferrule{}{\tentailsXt{\phi}{\eta}}}$~
+%\begin{mathpar}
+%\small
+%\inferrule{
+%	\tof{\fvar{fam}}{\eta} \in \Sigma
+%}{
+%	\tentailsXt{\fvar{fam}}{\eta}
+%}~(\text{var}_\phi)
+%
+%\inferrule{
+%	\kentailsX{\kFamStd}\\
+%	\tentailsXt{\psirep}{\karrow{\kappaidx}{\kIType}}\\
+%	\tentailsXt{\theta}{\Theta}
+%}{
+%	\tentailsXt{\tfamSpecStd}{\kFamStd}
+%}~(\text{family}_I)
+%\end{mathpar}
+%$\fbox{\inferrule{}{\tentailsXt{\theta}{\Theta}}}$
+%\begin{mathpar}
+%\small
+%\inferrule{ }{
+%	\tentailsXt{\cdot}{\cdot}
+%}~(\text{no-ops})
+%
+%\inferrule{
+%	\tentailst{\Sigma, \tof{\fvar{fam}}{\eta}~\Delta}{\theta}{\Theta}\\
+%	\tentailst{\Sigma, \tof{\fvar{fam}}{\eta}~\Delta}{\psi}{\karrow{\kType{\eta}}{\kappa}}
+%}{
+%	\tentailst{\Sigma, \tof{\fvar{fam}}{\eta}~\Delta}{\topp{\theta}{\tops{id}{\psi}}}{\Topp{\Theta}{\Tops{id}{\kappa}}}
+%}~(\text{op-seq})
+%\end{mathpar}
+%\caption{Kind checking rules for $\lamA$. All contexts have standard substructural properties (not shown).}
+%\end{figure*}
+%\begin{figure*}
+%$\fbox{\inferrule{}{\kentailsX{\kappa}}}$~
+%$\fbox{\inferrule{}{\kentailsX{\kSimple{\kappa}}}}$
+%%~~~~$\$
+%\begin{mathpar}
+%\small
+%\inferrule{
+%	\kentailsX{\kappa_1}\\
+%	\cdots\\
+%	\kentailsX{\kappa_n}\\
+%	\kentailsX{\kappa}
+%}{
+%	\kentailsX{\karrow{\splat{\kappa}{1}{n}}{\kappa}}
+%}~(\text{$\rightarrow$}_\kappa)
+%
+%\inferrule{
+%	\kentailsX{\kappa_1}\\
+%	\kentailsX{\kappa_2}
+%}{
+%	\kentailsX{\kpair{\kappa_1}{\kappa_2}}
+%}~(\text{$\times$}_\kappa)
+%
+%\inferrule{
+%	\kentailsX{\kSimple{\kappa}}
+%}{
+%	\kentailsX{\kappa}
+%}~(\text{simple})
+%
+%\inferrule{
+%	\kentailsX{\kSimple{\kappa_1}}\\
+%	\kentailsX{\kSimple{\kappa_2}}
+%}{
+%	\kentailsX{\kSimple{\kpair{\kappa_1}{\kappa_2}}}
+%}~(\text{$\times_\kappa$-simple})
+%
+%\inferrule{ }{
+%	\kentailsX{\kSimple{\kunit}}
+%}~(\text{Unit}_\kappa)
+%
+%\inferrule{ }{
+%	\kentailsX{\kSimple{\kTypeBlur}}
+%}~(\text{Type-$\subseteq$}_\kappa)
+%
+%\inferrule{
+%	\kentailsX{\eta}
+%}{
+%	\kentailsX{\kSimple{\kType{\eta}}}
+%}~(\text{Type}_\kappa)
+%
+%\inferrule{ }{
+%	\kentailsX{\kSimple{\kDen}}
+%}~(\text{Den-$\subseteq$}_\kappa)
+%
+%\inferrule{
+%	\kentailsX{\eta}
+%}{
+%	\kentailsX{\kSimple{\kDenk{\eta}}}
+%}~(\text{Den}_\kappa)
+%
+%\inferrule{ }{
+%	\kentailsX{\kSimple{\kIType}}
+%}~(\text{IType}_\kappa)
+%
+%\inferrule{ }{
+%	\kentailsX{\kSimple{\kITerm}}
+%}~(\text{ITerm}_\kappa)
+%
+%\inferrule{ }{
+%	\kentailsX{\kSimple{\kProg}}
+%}~(\text{Program}_\kappa)
+%
+%\inferrule{ }{
+%	\kentailsX{\kSimple{\kOp{n}}}
+%}~(\text{$n$-Op}_\kappa)
+%\end{mathpar}
+%$\fbox{\inferrule{}{\kentailsX{\eta}}}$~
+%\begin{mathpar}
+%\small
+%\inferrule{
+%	\kentailsX{\kSimple{\kappaidx}}\\
+%	\kentailsX{\Theta}
+%}{
+%	\kentailsX{\kFamStd}
+%}~(\text{family}_\eta)
+%\end{mathpar}
+%$\fbox{\inferrule{}{\kentailsX{\Theta}}}$
+%\begin{mathpar}
+%\small
+%\inferrule{~}{
+%	\kentailsX{\cdot}
+%}~(\text{No-Ops})
+%%
+%%\inferrule{
+%%	\kentailsX{\Theta}
+%%}{
+%%	\kentailsX{\Topp{\Theta}{\Tops{id}{\kOp{n}}}}
+%%}~(\text{(0, $n$)-Op-Seq})
+%
+%\inferrule{
+%	\kentailsX{\Theta}\\
+%	\kentailsX{\karrow{\splat{\kappa}{1}{m}}}{\kOp{n}}
+%}{
+%	\kentailsX{\Topp{\Theta}{\Tops{id}{\karrow{\splat{\kappa}{1}{m}}{\kOp{n}}}}}
+%}~(\text{($m$, $n$)-Op-Seq})
+%\end{mathpar}
+%$\fbox{\inferrule{}{\eentailsX{e}}}$
+%~~~~$\etvarCtx ::= \cdot \pipe \etvarCtx, \evar{x}$
+%\begin{mathpar}
+%\small
+%\inferrule{
+%	\evar{x} \in \etvarCtx
+%}{
+%	\eentailsX{\evar{x}}
+%}~(\text{var-form}_e)
+%
+%\inferrule{
+%	\tentailsXt{\psi}{\kTypeBlur}\\
+%	\eentails{\fCtx}{\tvarCtx}{\etvarCtx, \evar{x}}{e}
+%}{
+%	\eentailsX{\elam{\evar{x}}{\psi}{e}}
+%}~(\text{$\lambda$-form}_e)
+%
+%\inferrule{
+%	\eentailsX{e_1}\\
+%	\eentailsX{e_2}
+%}{
+%	\eentailsX{\eapp{e_1}{e_2}}
+%}~(\text{app-form}_e)
+%
+%\inferrule{
+%	\tentailsXt{\psitype}{\kType{\kFamStd}}\\
+%	\tof{\tvar{id}}{\karrow{\splat{\kappa}{1}{m}}{\kOp{0}}}\in\Theta\\
+%	\tentailsXt{\psi_1}{\kappa_1}\\
+%	\cdots\\
+%	\tentailsXt{\psi_m}{\kappa_m}
+%}{
+%	\eentailsX{\eop{\psitype}{\tvar{id}}{\splat{\psi}{1}{m}}{ }}
+%}~(\text{$(m, 0)$-op-app-form}_e)
+%
+%\inferrule{
+%	\tentailsXt{\psitype}{\kType{\kFamStd}}~~~~~~~~
+%	\tof{\tvar{id}}{
+%		\karrow{\splat{\kappa}{1}{m}}{
+%			\kOp{n}
+%		}
+%	} \in \Theta~~~~~~~~
+%	\tentailsXt{\psi_1}{\kappa_1}~~~~~~~~
+%	\cdots~~~~~~~~
+%	\tentailsXt{\psi_m}{\kappa_m}\\
+%	\eentailsX{e_1}\\
+%	\cdots\\
+%	\eentailsX{e_n}
+%}{
+%	\eentailsX{\eop{\psitype}{\tvar{id}}{\splat{\psi}{1}{m}}{\splat{e}{1}{n}}}
+%}~(\text{$(m, n)$-op-app-form}_e)
+%\end{mathpar}
+%$\fbox{\inferrule{}{\ientailsX{\mu}}}$
+%~~~~$\itvarCtx ::= \cdot \pipe \itvarCtx, \ivar{u}$
+%\begin{mathpar}
+%\small
+%\inferrule{
+%	\ivar{u} \in \itvarCtx
+%}{
+%	\ientailsX{\ivar{u}}
+%}~(\text{var-form}_{\mu})
+%
+%\inferrule{
+%	\tentailsX{\delta}\\
+%	\ientails{\iMkCtx{\fCtx}{\tvarCtx}{\itvarCtx}, \ivar{u}}{\mu}
+%}{
+%	\ientailsX{\ilam{\ivar{u}}{\delta}{\mu}}
+%}~(\text{$\lambda$-form}_{\mu})
+%
+%\inferrule{
+%	\tentailsX{\delta}\\
+%	\ientails{\iMkCtx{\fCtx}{\tvarCtx}{\itvarCtx}, \ivar{f}}{\mu}
+%}{
+%	\ientailsX{\ifix}{\ivar{f}}{\delta}{\mu}
+%}~(\text{fix-form}_\mu)
+%
+%\inferrule{
+%	\ientailsX{\mu_1}\\
+%	\ientailsX{\mu_2}
+%}{
+%	\ientailsX{\iapp}{\mu_1}{\mu_2}
+%}~(\text{app-form}_\mu)
+%
+%\inferrule{
+%	\ientailsX{\mu_1}\\
+%	\ientailsX{\mu_2}\\
+%	\ientailsX{\mu_3}\\
+%	\ientailsX{\mu_4}
+%}{
+%	\ientailsX{\iIfEq}{\mu_1}{\mu_2}{\mu_3}{\mu_4}
+%}~(\text{cond-form}_\mu)
+%
+%\inferrule{ }{\ientailsX{\iintlit}}~(\text{int-form}_\mu)
+%
+%\inferrule{
+%	\ientailsX{\mu_1}\\
+%	\ientailsX{\mu_2}
+%}{
+%	\ientailsX{\iop{\mu_1}{\mu_2}}
+%}~(\text{+-form}_\mu)
+%
+%\inferrule{
+%	\ientailsX{\mu_1}\\
+%	\ientailsX{\mu_2}
+%}{
+%	\ientailsX{\ipair{\mu_1}{\mu_2}}
+%}~(\text{$\times$-form}_\mu)
+%
+%\inferrule{
+%	\ientailsX{\mu}
+%}{
+%	\ientailsX{\ifst{\mu}}
+%}~(\text{fst-form}_\mu)
+%
+%\inferrule{
+%	\ientailsX{\mu}
+%}{
+%	\ientailsX{\isnd{\mu}}
+%}~(\text{snd-form}_\mu)
+%
+%\inferrule{
+%	\tentailsXt{\psi}{\kITerm}
+%}{
+%	\ientailsX{\iup{\psi}}
+%}~(\text{$\uparrow$-form}_\mu)
+%\end{mathpar}
+%$\fbox{\inferrule{}{\tentailsX{\delta}}}$
+%\begin{mathpar}
+%\small
+%\inferrule{
+%	\tentailsX{\delta_1}\\
+%	\tentailsX{\delta_2}
+%}{
+%	\tentailsX{\darrow{\delta_1}{\delta_2}}
+%}~(\text{$\rightarrow$-form}_\delta)
+%
+%\inferrule{ }{\tentailsX{\dint}}~(\text{int-form}_\delta)
+%
+%\inferrule{
+%	\tentailsX{\delta_1}\\
+%	\tentailsX{\delta_2}
+%}{
+%	\tentailsX{\dpair{\delta_1}{\delta_2}}
+%}~(\text{$\times$-form}_\delta)
+%
+%\inferrule{
+%	\tentailsXt{\psi}{\kIType}
+%}{
+%	\tentailsX{\dup{\psi}}
+%}~(\text{$\uparrow$-form}_\delta)
+%\end{mathpar}
+%\caption{Kind and term formation rules}
+%\end{figure*}
+%
+%\begin{figure*}
+%%  \renewcommand\thefigure{4}
+%% % \begin{textblock}{20}(0,0)
+%%  \newpage
+%$\fbox{$\tevalX{\psi}{\psi'}$}$~
+%$\fbox{$\tevalX{\theta}{\theta'}$}$
+%$\fbox{$\tevalX{\phi}{\phi'}$}$
+%~~~~$\famEvalCtx ::= \cdot \pipe \famEvalCtx,\tfamSpecStd$
+%~~~~~~~~$\errCtx ::= \noprob \pipe \prob \pipe \errCtx, \errCtx$
+%\begin{mathpar}
+%\small
+%\inferrule{ }{
+%	\tevalsX{\tunit}{\tunit}{\noprob}
+%}~(\text{unit}_\beta)
+%
+%\inferrule{
+%	\tevalsX{\psi_1}{\psi_1'}{\errCtx_{1}}\\
+%	\tevalsX{\psi_2}{\psi_2'}{\errCtx_{2}}
+%}{
+%	\tevalsX{\tpair{\psi_1}{\psi_2}}{\tpair{\psi_1'}{\psi_2'}}{\errCtx_{1},\errCtx_{2}}
+%}~(\text{pair}_\beta)
+%
+%\inferrule{
+%	\tevalX{\psi}{\tpair{\psi_1}{\psi_2}}
+%}{
+%	\tevalX{\tfst{\psi}}{\psi_1}
+%}~(\text{fst}_\beta)
+%
+%\inferrule{
+%	\tevalX{\psi}{\tpair{\psi_1}{\psi_2}}
+%}{
+%	\tevalX{\tsnd{\psi}}{\psi_2}
+%}~(\text{snd}_\beta)
+%
+%\inferrule{ }{
+%	\tevalsX{\tlam{\splatTwo{\tvar{t}}{{:}\kappa}{1}{n}}{\psi}}{\tlam{\splatTwo{\tvar{t}}{{:}\kappa}{1}{n}}{\psi}}{\noprob}
+%}~(\lambda_\beta)
+%
+%\inferrule{
+%	\tevalsX{\psi_0}{\tlam{\splatTwo{\tvar{t}}{{:}\kappa}{1}{n}}{\psi}}{\errCtx_{0}}\\
+%	\tevalsX{\psi_1}{\psi_1'}{\errCtx_{1}}\\
+%	\cdots\\
+%	\tevalsX{\psi_n}{\psi_n'}{\errCtx_{n}}\\
+%	\tevalsX{\substn{\splatTwo{\psi'}{{/}t}{1}{n}}{\psi}}{\psi'}{\errCtx}
+%}{
+%	\tevalsX{\tapp{\psi_0}{\splat{\psi}{1}{n}}}{\psi'}{\errCtx_{0},\errCtx_{1},\ldots,\errCtx_{n},\errCtx}
+%}~(\text{app}_\beta)
+%
+%\inferrule{
+%	\tevalsX{\psi_0}{\psi_0'}{\errCtx_{0}}\\
+%	\tevalsX{\psi_1}{\psi_0'}{\errCtx_{1}}\\
+%	\tevalsX{\psi_2}{\psi_2'}{\errCtx_{2}}
+%}{
+%	\tevalsX{\tifeq{\psi_0}{\psi_1}{\psi_2}{\psi_3}}{\psi_2'}{\errCtx_{0},\errCtx_{1},\errCtx_{2}}
+%}~(\text{if}_\beta^1)
+%
+%\inferrule{
+%	\tevalsX{\psi_0}{\psi_0'}{\errCtx_{0}}\\
+%	\tevalsX{\psi_1}{\psi_1'}{\errCtx_{1}}\\
+%	\psi_0' \neq \psi_1'\\
+%	\tevalsX{\psi_3}{\psi_3'}{\errCtx_{3}}
+%}{
+%	\tevalsX{\tifeq{\psi_0}{\psi_1}{\psi_2}{\psi_3}}{\psi_3'}{\errCtx_{0},\errCtx_{1},\errCtx_{3}}
+%}~(\text{if}_\beta^2)
+%
+%\inferrule{
+%	\tevalsX{\tfamSpecStd}{\phi}{\errCtx_0}\\
+%	\tevalsX{\subst{\phi}{\fvar{fam}}{\psi}}{\psi'}{\errCtx_1}
+%}{
+%	\tevalsX{\tfamStd}{\psi'}{\errCtx_0, \errCtx_1}
+%}~(\text{fambind}_\beta)
+%
+%\inferrule{
+%	\tevalsX{\psirep}{\psirep'}{\errCtx_0}\\
+%	\tevalsX{\theta}{\theta'}{\errCtx_1}
+%}{
+%	\tevalsX{\tfamSpecStd}{\tfamSpec{fam}{\kappaidx}{\psirep'}{\theta'}{\Theta}}{\errCtx_0, \errCtx_1}
+%}~(\text{family}_\beta)
+%
+%\inferrule{
+%	\tevalsX{\psitype}{\ttype{\psiidx}{\phi'}}{\errCtx_0}\\
+%	\tevalsX{\phi}{\phi'}{\errCtx_1}\\
+%	\tevalsX{\tapp{\psi_0}{\ttype{\psiidx}{\phi'}}}{\psi_0'}{\errCtx_2}
+%}{
+%	\tevalsX{
+%		\tfamcase{\psitype}{\phi}{\psi_0}{\psi_1}
+%	}{
+%		\psi_0'
+%	}{\errCtx_0, \errCtx_1, \errCtx_2}
+%}~(\text{famcase-T}_\beta^1)
+%
+%\inferrule{
+%	\tevalsX{\psitype}{\ttype{\psiidx}{\phi'}}{\errCtx_0}\\
+%	\tevalsX{\phi}{\phi''}{\errCtx_1}\\
+%	\phi' \neq \phi''\\
+%	\tevalsX{\psi_1}{\psi_1'}{\errCtx_2}
+%}{
+%	\tevalsX{
+%		\tfamcase{\psitype}{\phi}{\psi_0}{\psi_1}
+%	}{
+%		\psi_1'
+%	}{\errCtx_0, \errCtx_1, \errCtx_2}
+%}~(\text{famcase-T}_\beta^2)
+%
+%
+%\inferrule{
+%	\tevalsX{\psiden}{\tden{\psiiterm}{\ttype{\psiidx}{\phi'}}}{\errCtx_0}\\
+%	\tevalsX{\phi}{\phi'}{\errCtx_1}\\
+%	\tevalsX{\tapp{\psi_0}{\tden{\psiiterm}{\ttype{\psiidx}{\phi'}}}}{\psi_0'}{\errCtx_2}
+%}{
+%	\tevalsX{
+%		\tfamcase{\psiden}{\phi}{\psi_0}{\psi_1}
+%	}{
+%		\psi_0'
+%	}{\errCtx_0, \errCtx_1, \errCtx_2}
+%}~(\text{famcase-D}_\beta^1)
+%
+%\inferrule{
+%	\tevalsX{\psiden}{\tden{\psiiterm}{\ttype{\psiidx}{\phi'}}}{\errCtx_0}\\
+%	\tevalsX{\phi}{\phi''}{\errCtx_1}\\
+%	\phi' \neq \phi''\\
+%	\tevalsX{\psi_1}{\psi_1'}{\errCtx_2}
+%}{
+%	\tevalsX{
+%		\tfamcase{\psiden}{\phi}{\psi_0}{\psi_1}
+%	}{
+%		\psi_1'
+%	}{\errCtx_0, \errCtx_1, \errCtx_2}
+%}~(\text{famcase-D}_\beta^2)
+%
+%\inferrule{
+%	\tevalsX{\psiidx}{\psiidx'}{\errCtx}
+%}{
+%	\tevalsX{\ttypestd}{
+%		\ttype{\psiidx'}{fam}
+%	}{\errCtx}
+%}
+%
+%\inferrule{
+%	\tevalsX{\psitype}{\ttypestd}{\errCtx}
+%}{
+%	\tevalsX{\tidx{\psitype}}{\psiidx}{\errCtx}
+%}
+%
+%\inferrule{
+%	\tevalsX{\psitype}{\ttypestd}{\errCtx_0}\\
+%	\tfamSpecStd \in \famEvalCtx\\
+%	\tevalsX{\tapp{\psirep}{\psiidx}}{\psi'}{\errCtx_1}
+%}{
+%	\tevalsX{\trepof{\psitype}}{\psi'}{\errCtx_0, \errCtx_1}
+%}
+%
+%\inferrule{
+%	\tevalsX{\psi}{\psi'}{\errCtx}
+%}{
+%	\tevalsX{\tconst{\psi}}{\tconst{\psi'}}{\errCtx}
+%}
+%
+%\inferrule{
+%	\tevalsX{\psi}{\psi'}{\errCtx}
+%}{
+%	\tevalsX{\tOp{\psi}}{\tOp{\psi'}}{\errCtx}
+%}
+%
+%\inferrule{
+%	\tevalsX{\psiiterm}{\psiiterm'}{\errCtx_0}\\
+%	\tevalsX{\psitype}{\psitype'}{\errCtx_1}
+%}{
+%	\tevalsX{\tden{\psiiterm}{\psitype}}{\tden{\psiiterm'}{\psitype'}}{\errCtx_0, \errCtx_1}
+%}
+%
+%\inferrule{
+%	\tevalsX{\psiden}{\tden{\psiiterm}{\psitype}}{\errCtx}
+%}{
+%	\tevalsX{\tvalof{\psiden}}{\psiiterm}{\errCtx}
+%}
+%
+%\inferrule{
+%	\tevalsX{\psiden}{\tden{\psiiterm}{\psitype}}{\errCtx}
+%}{
+%	\tevalsX{\ttypeof{\psiden}}{\psitype}{\errCtx}
+%}
+%
+%\inferrule{ }{
+%	\tevalsX{\terr}{\terr}{\noprob}
+%}
+%
+%\inferrule{
+%	\tevalsX{\mu}{\mu'}{\errCtx}
+%}{
+%	\tevalsX{\titerm{\mu}}{\titerm{\mu'}}{\errCtx}
+%}
+%
+%\inferrule{
+%	\tevalsX{\delta}{\delta'}{\errCtx}
+%}{
+%	\tevalsX{\titype{\delta}}{\titype{\delta'}}{\errCtx}
+%}
+%
+%\inferrule{
+%	\tevale{e}{e'}
+%}{
+%	\tevalX{\tprog{e}}{\tprog{e'}}
+%}
+%
+%\inferrule{ }{
+%	\tevalsX{\topsempty}{\topsempty}{\noprob}
+%}
+%
+%\inferrule{
+%	\tevalsX{\theta}{\theta'}{\errCtx_0}\\
+%	\tevalsX{\psi}{\psi'}{\errCtx_1}
+%}{
+%	\tevalsX{
+%		\topp{\theta}{\tops{op}{\psi}}
+%	}{
+%		\topp{\theta'}{\tops{op}{\psi'}}
+%	}{\errCtx_0, \errCtx_1}
+%}
+%\end{mathpar}
+%$\fbox{$\tevale{e}{e'}$}$
+%\begin{mathpar}
+%\small
+%\inferrule{
+%	\tevalsX{\psi}{\psi'}{\errCtx}
+%}{
+%	\tevalsX{\elam{\evar{x}}{\psi}{e}}
+%	{\elam{\evar{x}}{\psi'}{e}}{\errCtx}
+%}
+%
+%\inferrule{
+%	\tevalsX{e_1}{e_1'}{\errCtx_1}\\
+%	\tevalsX{e_2}{e_2'}{\errCtx_2}
+%}{
+%	\tevalsX{\eapp{e_1}{e_2}}{\eapp{e_1'}{e_2'}}{\errCtx_1, \errCtx_2}
+%}
+%
+%\inferrule{
+%	\tevalsX{\psitype}{\psitype'}{\errCtx}\\
+%	\tevalsX{\psi_1}{\psi_1'}{\errCtx_1^1}\\
+%	\cdots\\
+%	\tevalsX{\psi_m}{\psi_m'}{\errCtx_m^1}\\
+%	\tevalsX{e_1}{e_1'}{\errCtx_1^2}\\
+%	\cdots\\
+%	\tevalsX{e_n}{e_n'}{\errCtx_n^2}
+%}{
+%	\tevalsX{\eop{\psitype}{op}{\splat{\psi}{1}{m}}{\splat{e}{1}{n}}}
+%	{\eop{\psitype'}{op}{\splat{\psi'}{1}{m}}{\splat{e'}{1}{n}}}
+%	{\errCtx, \errCtx_1^1, \ldots, \errCtx_m^1, \errCtx_1^2, \ldots, \errCtx_n^2}
+%}
+%\end{mathpar}
+%$\fbox{$\tevalm{\mu}{\mu'}$}$
+%\begin{mathpar}
+%\small
+%\inferrule{
+%	\tevalsX{\delta}{\delta'}{\errCtx}
+%}{
+%	\tevalsX{\ilam{\ivar{u}}{\delta}{\mu}}{\ilam{\ivar{u}}{\delta'}{\mu}}{\errCtx}
+%}
+%
+%\inferrule{
+%	\tevalsX{\delta}{\delta'}{\errCtx}
+%}{
+%	\tevalsX{\ifix{\ivar{f}}{\delta}{\mu}}{\ifix{\ivar{f}}{\delta'}{\mu}}{\errCtx}
+%}
+%
+%\inferrule{
+%	\tevalsX{\mu_1}{\mu_1'}{\errCtx_1}\\
+%	\tevalsX{\mu_2}{\mu_2'}{\errCtx_2}
+%}{
+%	\tevalsX{\iapp{\mu_1}{\mu_2}}{\iapp{\mu_1'}{\mu_2'}}{\errCtx_1, \errCtx_2}
+%}
+%
+%\inferrule{
+%	\tevalsX{\mu_0}{\mu_0'}{\errCtx_0}\\
+%	\tevalsX{\mu_1}{\mu_1'}{\errCtx_1}\\
+%	\tevalsX{\mu_2}{\mu_2'}{\errCtx_2}\\
+%	\tevalsX{\mu_3}{\mu_3'}{\errCtx_3}
+%}{
+%	\tevalsX{\iIfEq{\mu_0}{\mu_1}{\mu_2}{\mu_3}}{\iIfEq{\mu_0'}{\mu_1'}{\mu_2'}{\mu_3'}}{\errCtx_0, \errCtx_1, \errCtx_2, \errCtx_3}
+%}
+%
+%\inferrule{ }{
+%	\tevalsX{\iintlit}{\iintlit}{\noprob}
+%}
+%
+%\inferrule{
+%	\tevalsX{\mu_1}{\mu_1'}{\errCtx_1}\\
+%	\tevalsX{\mu_2}{\mu_2'}{\errCtx_2}
+%}{
+%	\tevalsX{\iop{\mu_1}{\mu_2}}{\iop{\mu_1'}{\mu_2'}}{\errCtx_1, \errCtx_2}
+%}
+%
+%\inferrule{
+%	\tevalsX{\mu_1}{\mu_1'}{\errCtx_1}\\
+%	\tevalsX{\mu_2}{\mu_2'}{\errCtx_2}
+%}{
+%	\tevalsX{\ipair{\mu_1}{\mu_2}}{\ipair{\mu_1'}{\mu_2'}}{\errCtx_1, \errCtx_2}
+%}
+%
+%\inferrule{
+%	\tevalsX{\mu}{\mu'}{\errCtx}
+%}{
+%	\tevalsX{\ifst{\mu}}{\ifst{\mu'}}{\errCtx}
+%}
+%
+%\inferrule{
+%	\tevalsX{\mu}{\mu'}{\errCtx}
+%}{
+%	\tevalsX{\isnd{\mu}}{\isnd{\mu'}}{\errCtx}
+%}
+%
+%\inferrule{
+%	\tevalsX{\psi}{\titerm{\mu}}{\errCtx}
+%}{
+%	\tevalsX{\iup{\psi}}{\mu}{\errCtx}
+%}
+%\end{mathpar}
+%$\fbox{$\tevalm{\delta}{\delta'}$}$
+%\begin{mathpar}
+%\small
+%\inferrule{
+%	\tevalsX{\delta_1}{\delta_1'}{\errCtx_1}\\
+%	\tevalsX{\delta_2}{\delta_2'}{\errCtx_2}
+%}{
+%	\tevalsX{\darrow{\delta_1}{\delta_2}}{\darrow{\delta_1'}{\delta_2'}}{\errCtx_1, \errCtx_2}
+%}
+%
+%\inferrule{ }{
+%	\tevalsX{\dint}{\dint}{\noprob}
+%}
+%
+%\inferrule{
+%	\tevalsX{\delta_1}{\delta_1'}{\errCtx_1}\\
+%	\tevalsX{\delta_2}{\delta_2'}{\errCtx_2}
+%}{
+%	\tevalsX{\dpair{\delta_1}{\delta_2}}{\dpair{\delta_1'}{\delta_2'}}{\errCtx_1, \errCtx_2}
+%}
+%
+%\inferrule{
+%	\tevalsX{\psi}{\titype{\delta}}{\errCtx}
+%}{
+%	\tevalsX{\dup{\psi}}{\delta}{\errCtx}
+%}
+%\end{mathpar}
+%%
+%%\inferrule{
+%%	\tevalX{\psi}{\ttype{\psiidx}{\kappa}{\psirep}}
+%%}{
+%%	\tevalX{\tidx{\psi}}{\psiidx}
+%%}
+%%
+%%\inferrule{
+%%	\tevalX{\psi}{\ttype{\psiidx}{\kappa}{\psirep}}
+%%}{
+%%	\tevalX{\trepof{\psi}}{\psirep}
+%%}
+%%
+%%\inferrule{
+%%	\tevalsX{\psi}{\ttype{\psiidx}{\kappa}{\psirep}}{\errCtx_{0}}\\
+%%	\tevalsX{\tapp{\psi_{0}}{\ttype{\psiidx}{\kappa}{\psirep}}}{\psi'}{\errCtx_{1}}
+%%}{
+%%	\tevalsX{\tfamcase{\psi}{\kappa}{\psi_{0}}{\psi_{1}}}{\psi'}{\errCtx_{0},\errCtx_{1}}
+%%}
+%%
+%%\inferrule{
+%%	\tevalsX{\psi}{\ttype{\psiidx}{\kappa'}{\psirep}}{\errCtx_{0}}\\
+%%	\kappa \neq \kappa'\\
+%%	\tevalsX{\psi_{1}}{\psi'}{\errCtx_{1}}
+%%}{
+%%	\tevalsX{\tfamcase{\psi}{\kappa}{\psi_{0}}{\psi_{1}}}{\psi'}{\errCtx_{0},\errCtx_{1}}
+%%}
+%%
+%%\inferrule{
+%%	\tevalX{\psi}{\psi'}
+%%}{
+%%	\tevalX{\tconst{\psi}}{\tconst{\psi'}}
+%%}
+%%
+%%\inferrule{
+%%	\tevalX{\psi}{\psi'}
+%%}{
+%%	\tevalX{\tOp{\psi}}{\tOp{\psi'}}
+%%}
+%%
+%%\inferrule{
+%%	\tevalms{\mu}{\mu'}{\errCtx_{0}}\\
+%%	\tevalsX{\psitype}{\ttype{\psiidx}{\kappa}{\psirep}}{\errCtx_{1}}\\
+%%	\mtentails{\cdot}{\mtof{\mu'}{\psirep}}
+%%}{
+%%	\tevalsX{\tden{\mu}{\psitype}}{\tden{\mu'}{\psitype'}}{\errCtx_{0},\errCtx_{1}}
+%%}
+%%
+%%\inferrule{
+%%	\tevalms{\mu}{\mu'}{\errCtx_{0}}\\
+%%	\tevalsX{\psitype}{\ttype{\psiidx}{\kappa}{\psirep}}{\errCtx_{1}}\\
+%%	\cdot \nvdash \mtof{\mu'}{\psirep}
+%%}{
+%%	\tevalsX{\tden{\mu}{\psitype}}{\tden{\mu'}{\psitype'}}{\errCtx_{0},\errCtx_{1},\prob}
+%%}
+%%
+%%\inferrule{
+%%	\tevalX{\psi}{\tden{\mu}{\psitype}}
+%%}{
+%%	\tevalX{\ttypeof{\psi}}{\psitype}
+%%}
+%%
+%%\inferrule{ }{
+%%	\tevalsX{\terr}{\terr}{\noprob}
+%%}
+%%
+%%\inferrule{
+%%	\tevalsX{\psi}{\tden{\mu}{\ttype{\psiidx}{\kappa}{\psirep}}}{\errCtx_{0}}\\
+%%	\tevalsX{\tapp{\psi_{0}}{\tden{\mu}{\ttype{\psiidx}{\kappa}{\psirep}}}}{\psi'}{\errCtx_{1}}
+%%}{
+%%	\tevalsX{\ttidxcase{\psi}{\kappa}{\psi_{0}}{\psi_{1}}}{\psi'}{\errCtx_{0},\errCtx_{1}}
+%%}
+%%
+%%\inferrule{
+%%	\tevalsX{\psi}{\terr}{\errCtx_{0}}\\
+%%	\tevalsX{\psi_{1}}{\psi'}{\errCtx_{1}}
+%%}{
+%%	\tevalsX{\ttidxcase{\psi}{\kappa}{\psi_{0}}{\psi_{1}}}{\psi'}{\errCtx_{0},\errCtx_{1}}
+%%}
+%%
+%%\inferrule{
+%%	\tevalsX{\psi}{\tden{\mu}{\ttype{\psiidx}{\kappa'}{\psirep}}}{\errCtx_{0}}\\
+%%	\kappa \neq \kappa'\\
+%%	\tevalsX{\psi_{1}}{\psi'}{\errCtx_{1}}
+%%}{
+%%	\tevalsX{\ttidxcase{\psi}{\kappa}{\psi_{0}}{\psi_{1}}}{\psi'}{\errCtx_{0},\errCtx_{1}}
+%%}
+%%
+%%\inferrule{
+%%	\tevalsX{\psi_1}{\psi_1'}{\errCtx_{0}}\\
+%%	\tevalsX{\psi_2}{\psi_2'}{\errCtx_{1}}
+%%}{
+%%	\tevalsX{\tILtarrow{\psi_1}{\psi_2}}{\tILtarrow{\psi_1'}{\psi_2'}}{\errCtx_{0},\errCtx_{1}}
+%%}
+%%
+%%\inferrule{
+%%	\tevalsX{\psi_1}{\psi_1'}{\errCtx_{0}}\\
+%%	\tevalsX{\psi_2}{\psi_2'}{\errCtx_{1}}
+%%}{
+%%	\tevalsX{\tILtprod{\psi_1}{\psi_2}}{\tILtprod{\psi_1'}{\psi_2'}}{\errCtx_{0},\errCtx_{1}}
+%%}
+%%
+%%\inferrule{ }{
+%%	\tevalsX{\tILtint}{\tILtint}{\noprob}
+%%}
+%%
+%%\inferrule{
+%%	\tevalsX{\psi}{\tILtarrow{\psi_1}{\psi_2}}{\errCtx_{0}}\\
+%%	\tevalsX{\tapp{\psiarrow}{\psi_1, \psi_2}}{\psi'}{\errCtx_{1}}
+%%}{
+%%	\tevalsX{\tILtcase{\psi}{\psiarrow}{\psiprod}{\psiint}}{\psi'}{\errCtx_{0},\errCtx_{1}}
+%%}
+%%
+%%\inferrule{
+%%	\tevalsX{\psi}{\tILtprod{\psi_1}{\psi_2}}{\errCtx_{0}}\\
+%%	\tevalsX{\tapp{\psiprod}{\psi_1, \psi_2}}{\psi'}{\errCtx_{1}}
+%%}{
+%%	\tevalsX{\tILtcase{\psi}{\psiarrow}{\psiprod}{\psiint}}{\psi'}{\errCtx_{0},\errCtx_{1}}
+%%}
+%%
+%%\inferrule{
+%%	\tevalsX{\psi}{\tILtint}{\errCtx_{0}}\\
+%%	\tevalsX{\psiint}{\psiint'}{\errCtx_{1}}
+%%}{
+%%	\tevalsX{\tILtcase{\psi}{\psiarrow}{\psiprod}{\psiint}}{\psiint'}{\errCtx_{0},\errCtx_{1}}
+%%}
+%%
+%%\inferrule{ }{
+%%	\tevalms{\ivar{u}}{\ivar{u}}{\noprob}
+%%}
+%%
+%%\inferrule{
+%%	\tevalX{\psiden}{\tden{\mu}{\psitype}}
+%%}{
+%%	\tevalm{\mvalof{\psiden}}{\mu}
+%%}
+%%
+%%\inferrule{
+%%	\tevalsX{\psi}{\psi'}{\errCtx_{0}}\\
+%%	\tevalms{\mu}{\mu'}{\errCtx_{1}}
+%%}{
+%%	\tevalms{\ilam{\ivar{u}}{\psi}{\mu}}{\ilam{\ivar{u}}{\psi'}{\mu'}}{\errCtx_{0},\errCtx_{1}}
+%%}
+%%
+%%\inferrule{
+%%	\tevalsX{\psi}{\psi'}{\errCtx_{0}}\\
+%%	\tevalms{\mu}{\mu'}{\errCtx_{1}}
+%%}{
+%%	\tevalms{\ifix{\ivar{f}}{\psi}{\mu}}{\ifix{\ivar{f}}{\psi'}{\mu'}}{\errCtx_{0},\errCtx_{1}}
+%%}
+%%
+%%\inferrule{
+%%	\tevalms{\mu_1}{\mu_1'}{\errCtx_{0}}\\
+%%	\tevalms{\mu_2}{\mu_2'}{\errCtx_{1}}
+%%}{
+%%	\tevalms{\iapp{\mu_1}{\mu_2}}{\iapp{\mu_1'}{\mu_2'}}{\errCtx_{0},\errCtx_{1}}
+%%}
+%%
+%%\inferrule{ }{
+%%	\tevalms{\iintlit}{\iintlit}{\noprob}
+%%}
+%%
+%%\inferrule{
+%%	\tevalms{\mu_1}{\mu_1'}{\errCtx_{0}}\\
+%%	\tevalms{\mu_2}{\mu_2'}{\errCtx_{1}}
+%%}{
+%%	\tevalms{\iop{\mu_1}{\mu_2}}{\iop{\mu_1'}{\mu_2'}}{\errCtx_{0},\errCtx_{1}}
+%%}
+%%
+%%\inferrule{
+%%	\tevalms{\mu_1}{\mu_1'}{\errCtx_{0}}\\
+%%	\tevalms{\mu_2}{\mu_2'}{\errCtx_{1}}
+%%}{
+%%	\tevalms{\ipair{\mu_1}{\mu_2}}{\ipair{\mu_1'}{\mu_2'}}{\errCtx_{0},\errCtx_{1}}
+%%}
+%%
+%%\inferrule{
+%%	\tevalm{\mu}{\mu'}
+%%}{
+%%	\tevalm{\ifst{\mu}}{\ifst{\mu'}}
+%%}
+%%
+%%\inferrule{
+%%	\tevalm{\mu}{\mu'}
+%%}{
+%%	\tevalm{\isnd{\mu}}{\isnd{\mu'}}
+%%}
+%%
+%%\inferrule{
+%%	\tevalms{\mu_0}{\mu_0'}{\errCtx_{0}}\\
+%%	\tevalms{\mu_1}{\mu_1'}{\errCtx_{1}}\\
+%%	\tevalms{\mu_2}{\mu_2'}{\errCtx_{2}}\\
+%%	\tevalms{\mu_3}{\mu_3'}{\errCtx_{3}}
+%%}{
+%%	\tevalms{\iIfEq{\mu_0}{\mu_1}{\mu_2}{\mu_3}}{\iIfEq{\mu_0'}{\mu_1'}{\mu_2'}{\mu_3'}}{\errCtx_{0},\errCtx_{1},\errCtx_{2},\errCtx_{3}}
+%%}
+%%
+%%\inferrule{ }{
+%%	\tevales{\evar{x}}{\evar{x}}{\noprob}
+%%}
+%%
+%%\inferrule{
+%%	\tevalsX{\psi}{\psi'}{\errCtx_{0}}\\
+%%	\tevales{e}{e'}{\errCtx_{1}}
+%%}{
+%%	\tevales{\elam{\evar{x}}{\psi}{e}}{\elam{\evar{x}}{\psi'}{e'}}{\errCtx_{0},\errCtx_{1}}
+%%}
+%%
+%%\inferrule{
+%%	\tevales{e_1}{e_1'}{\errCtx_{0}}\\
+%%	\tevales{e_2}{e_2'}{\errCtx_{1}}
+%%}{
+%%	\tevales{\eapp{e_1}{e_2}}{\eapp{e_1'}{e_2'}}{\errCtx_{0},\errCtx_{1}}
+%%}
+%%
+%%\inferrule{
+%%	\tevalX{\psi}{\psi'}
+%%}{
+%%	\tevale{\econst{\psi}}{\econst{\psi'}}
+%%}
+%%
+%%\inferrule{
+%%	\tevalsX{\psi}{\psi'}{\errCtx_{0}}\\
+%%	\tevales{e_1}{e_1'}{\errCtx_{1}}\\
+%%	\cdots\\
+%%	\tevales{e_n}{e_n'}{\errCtx_{n}}
+%%}{
+%%	\tevales{\eop{\psi}{\splat{e}{1}{n}}}{\eop{\psi'}{\splat{e'}{1}{n}}}{\errCtx_{0},\errCtx_{1},\ldots,\errCtx_{n}}
+%%}
+%\caption{Big-step operational semantics for type-level terms. The error context $\Sigma$ is used to detect representational consistency errors.}
+%%%\end{textblock}
+%\end{figure*}
+%%\begin{figure*}
+%%\renewcommand\thefigure{5}
+%%$\fbox{$\translates{\eCtx}{e}{\mu}{\psiidx}{\kappa}{\psirep}$}$
+%%~~~~$\eCtx ::= \cdot \pipe \extendeCtx{x}{\mu}{\psiidx}{\kappa}{\psirep}$
+%%\begin{mathpar}
+%%\inferrule{
+%%	\translatesTo{\evar{x}}{\mu}{\psiidx}{\kappa}{\psirep} \in \eCtx
+%%}{
+%%	\translatesX{\evar{x}}{\mu}{\psiidx}{\kappa}{\psirep}
+%%}
+%%
+%%\inferrule{ }{
+%%	\translatesX{
+%%		\econst{\tconst{\tden{\mu}{\ttype{\psiidx}{\kappa}{\psirep}}}}
+%%	}{
+%%		\mu
+%%	}{
+%%		\psiidx
+%%	}{
+%%		\kappa
+%%	}{
+%%		\psirep
+%%	}
+%%}
+%%
+%%\inferrule{
+%%	\translates{\extendeCtx{x}{\ivar{x}}{\psiidx}{\kappa}{\psirep}}{e}{\mu'}{\psiidx'}{\kappa'}{\psirep'}
+%%}{
+%%	\translatesX{
+%%		\elam{\evar{x}}{\ttype{\psiidx}{\kappa}{\psirep}}{e}
+%%	}{
+%%		\ilam{\ivar{x}}{\psirep}{\mu'}
+%%	}{
+%%		\tpair{\ttype{\psiidx}{\kappa}{\psirep}}{\ttype{\psiidx'}{\kappa'}{\psirep'}}
+%%	}{
+%%		\kpair{\kType{\kappa}}{\kType{\kappa'}}
+%%	}{
+%%		\tILtarrow{\psirep}{\psirep'}
+%%	}
+%%}
+%%
+%%\inferrule{
+%%	\translatesX{e_1}{\mu_1}{\psiidxn{1}}{\kappa_1}{\psirepn{1}}~~~~
+%%	\cdots~~~~
+%%	\translatesX{e_n}{\mu_n}{\psiidxn{n}}{\kappa_n}{\psirepn{n}}\\
+%%	\tevalsX{\tapp{\psi}{
+%%		\tden{\mu_1}{\ttype{\psiidxn{1}}{\kappa_1}{\psirepn{1}}},
+%%		\ldots,
+%%		\tden{\mu_n}{\ttype{\psiidxn{n}}{\kappa_n}{\psirepn{n}}}
+%%	}}{\tden{\mu}{\ttype{\psiidx}{\kappa}{\psirep}}}{\noprob}
+%%}{
+%%	\translatesX{
+%%		\eop{\tOp{\psi}}{\splat{e}{1}{n}}
+%%	}{
+%%		\mu
+%%	}{
+%%		\psiidx
+%%	}{
+%%		\kappa
+%%	}{
+%%		\psirep
+%%	}
+%%}
+%%
+%%\end{mathpar}
+%%\caption{Verification and translation rules for $\lamA$.}
+%%\end{figure*}
+%%\begin{figure*}
+%%  \renewcommand\thefigure{6}
+%%$\fbox{$\mtentailsXt{\mu}{\psi}$}$
+%%~~~~$\iCtx ::= \cdot \pipe \iCtx, \mtof{\ivar{u}}{\psi}$
+%%\begin{mathpar}
+%%\inferrule{
+%%	\mtof{\ivar{u}}{\psi} \in \ivarCtx
+%%}{
+%%	\mtentailsXt{\ivar{u}}{\psi}
+%%}
+%%
+%%\inferrule{
+%%	\mtentails{\iCtx, \mtof{\ivar{u}}{\psi}}{\mtof{\mu}{\psi'}}
+%%}{
+%%	\mtentailsXt{\ilam{\ivar{u}}{\psi}{\mu}}{\psi'}
+%%}
+%%
+%%\inferrule{
+%%	\mtentails{\iCtx, \mtof{\ivar{f}}{\psi}}{\mtof{\mu}{\psi}}
+%%}{
+%%	\mtentailsXt{\ifix{\ivar{f}}{\psi}{\mu}}{\psi}
+%%}
+%%
+%%\inferrule{
+%%	\mtentailsXt{\mu_1}{\tILtarrow{\psi_1}{\psi_2}}\\
+%%	\mtentailsXt{\mu_2}{\psi_1}
+%%}{
+%%	\mtentailsXt{\iapp{\mu_1}{\mu_2}}{\psi_2}
+%%}
+%%
+%%\inferrule{ }{
+%%	\mtentailsXt{\iintlit}{\tILtint}
+%%}
+%%
+%%\inferrule{
+%%	\mtentailsXt{\mu_1}{\tILtint}\\
+%%	\mtentailsXt{\mu_2}{\tILtint}
+%%}{
+%%	\mtentailsXt{\iop{\mu_1}{\mu_2}}{\tILtint}
+%%}
+%%
+%%\inferrule{
+%%	\mtentailsXt{\mu_1}{\psi_1}\\
+%%	\mtentailsXt{\mu_2}{\psi_2}
+%%}{
+%%	\mtentailsXt{\ipair{\mu_1}{\mu_2}}{\tILtprod{\psi_1}{\psi_2}}
+%%}
+%%
+%%\inferrule{
+%%	\mtentailsXt{\mu}{\tILtprod{\psi_1}{\psi_2}}
+%%}{
+%%	\mtentailsXt{\ifst{\mu}}{\psi_1}
+%%}
+%%
+%%\inferrule{
+%%	\mtentailsXt{\mu}{\tILtprod{\psi_1}{\psi_2}}
+%%}{
+%%	\mtentailsXt{\isnd{\mu}}{\psi_2}
+%%}
+%%
+%%\inferrule{
+%%	\mtentailsXt{\mu_0}{\psi_1}\\
+%%	\mtentailsXt{\mu_1}{\psi_1}\\
+%%	\mtentailsXt{\mu_2}{\psi_2}\\
+%%	\mtentailsXt{\mu_3}{\psi_2}
+%%}{
+%%	\mtentailsXt{\iIfEq{\mu_0}{\mu_1}{\mu_2}{\mu_3}}{\psi_2}
+%%}
+%%\end{mathpar}
+%%\caption{The internal type system of $\lamA$.}
+%%\end{figure*}
+%
+%%Let us now examine the definition of $\lamA$ in more detail.
+The grammar of $\lamA$ is given in Figure 2.
+Expressions in $\lamA$ can take one of only five forms, the first three of which correspond to the simply-typed lambda calculus and the remaining two to abstract constants and operator invocations, respectively.
+%%
+%%The type-level specification language is based on System F \cite{tapl} equipped with several additional primitives that we will describe below.
+To emphasize the expanded role for type-level terms, we use the letter $\psi$ instead of the more conventional $\tau$.
+It is constrained by a kind system, the rules of which are given in Figure 4.
+A big-step semantics for the type-level language is given in Figure 5.
+As we will state more formally below, well-kinded terms of the type-level language strongly normalize.
+%%
+%%Expressions are given meaning in a denotational style in terms of a typed internal language according to the verification and translation rules in Figure 6.
+The type system of the internal language is given in Figure 7.
+Unlike the type-level language, the internal language contains a ${\sf fix}$ operator, and is thus Turing-complete (corresponding to a variant of Plotkin's PCF \cite{tapl}.)
+%%
+%%\subsubsection{Central Compilation Judgement}
+%%The central compilation judgement of Figure 3 takes a type-level term that reduces to a program and produces a typed denotation for the program if and only if it is valid with respect to the kind system of $\lamA$ {\it and} the type system introduced by the user, the key feature of this mechanism.
+It consists of three steps:
+%%\begin{enumerate}
+%%\item \textbf{Kind checking}, where basic correctness properties of the user specification as well as uses of type-level terms in expressions are checked.
+%%\item \textbf{Specification Normalization}, which reduces the specification to normal form by internalizing relevant portions of it directly into the expressions of the program.
+%%\item \textbf{Verification and Translation}, where operator definitions are invoked to assign final denotations to expressions if possible, or derive that there is an $\terr$ if not.
+%%\end{enumerate}
+%%The rules given in Figures 4-7 are sufficient to define the central compilation judgement. We examine its theoretical properties in Section 2.4.
+%%
+%%\subsubsection{Type Constructors}
+%%
+%%An interesting feature of $\lamA$ is that there is just one form of type constructor: 
+%%$\ttype{\psiidx}{\kappa}{\psirep}$.
+%%All types are identified by their index, $\psiidx$ and an associated internal type, $\psirep$, which controls the representation of values that inhabit the type.
+The principal kind of a type constructor indexed by a value of kind $\kappa$ is $\kType{\kappa}$.
+%%
+%%
+%
+%afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf
+%
+%afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf
+%
+%afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf
+%
+%afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf
+%
+%afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf
+%
+%afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf
+%
+%afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf
+%
+%afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf
+%
+%afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf
+%
+%afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf
+%
+%afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf
+%
+%afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf
+%
+%afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf
+%
+%afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf
+%
+%afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf
+%
+%afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf
+%
+%afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf
+%
+%afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf
+%
+%afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf
+%
+%afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf
+%
+%afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf
+%
+%afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf
+%
+%afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf
+%
+%afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf
+%
+%afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf
+%
+%afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf afdasdfsdf
+%
+%
+%%In the example in Figure 1, $\tvar{nat}$ is a type with no (that is, unit) index that we will represent by an integer in the internal language.
+To distinguish it from other types of the same basic form, we need to label it.
+That is the purpose of the \textsf{label} declaration preceding the type definition -- it introduces a new kind of unit labeled $\klabel{nat}$.
+We can then identify the type $\tvar{nat}$ as being the sole type indexed by a value of kind $\klabel{nat}$ and represented by the internal type $\tILtint$.
+It is important to note that although the type $\tvar{nat}$ has kind $\kType{\klabel{nat}}$, it does not uniquely inhabit that kind -- you can construct other types indexed by a $\klabel{nat}$-labeled unit but associated with different internal types, so operator definitions must test for equality as appropriate.
+%%
+%%\subsubsection{Internal Terms and Types}
+%%The internal language, $\mu$, is a variant of Plotkin's PCF with integers, products and a conditional checks for $\beta$-equality.
+The form $\mvalof{\psi}$ is used to compose denotations, which are internal language terms paired with a type constructor (see below).
+In the type-level operational semantics, the $\mvalof{\psi}$ form  always reduces to one of the other forms of the internal language.
+%%
+%%Internal types are represented at the type-level by the constructors \texttt{arrow}, \texttt{prod} and \texttt{int}, which have kind $\kIType$.
+The internal type system is given in Figure 7.
+The language is equipped with the standard type safety theorem based on preservation and progress lemmas.
+Because expressions are given their semantics ultimately in terms of internal language terms, these global safety theorems are inherited by the language as a whole.
+In a more realistic language, additional desirable global safety restrictions could also be imposed using the internal type system (e.g.
+memory safety).
+%%
+%%\subsubsection{Function Types}
+%%It is worth considering now what type to assign to expression-level functions, e.g.
+$\elam{\evar{x}}{\tvar{nat}}{\eop{\tvar{s}}{\eop{\tvar{s}}{x}}}$.
+Because we have only one form for types, we must encode arrow types using this form.
+The encoding, given formally in Figure 6, encodes the arrow type $\psi_{1} \rightarrow \psi_{2}$, where $\psi_{1,2}$ are types indexed by values of kind $\kappa_{1,2}$, as
+%%$$\ttype{\tpair{\psi_{1}}{\psi_{2}}}{\kpair{\kType{\kappa_{1}}}{\kType{\kappa_{2}}}}{\tILtarrow{\trepof{\psi_{1}}}{\trepof{\psi_{2}}}}$$
+%%\noindent
+%%That is, the arrow type is a type indexed by a pair containing the constituent types and represented by the internal arrow type.
+As can be seen in the compilation rules, expression-level functions translate directly into internal language functions.
+%%\subsubsection{Operators}
+%%In order to construct and manipulate terms of a newly introduced type, we must provide primitive operators, introduced using the forms \textsf{const} and \textsf{op}.
+The kind associated with operators is $\kOp{n}$, where $n$ is the operator's arity.
+An operator must have fixed arity.
+Constants can be thought of as nullary operators.
+%%
+%%Constants are given meaning directly using a denotation.
+The constant $\tvar{z}$, for example, has the denotation $\tden{0}{\tvar{nat}}$, reflecting the fact that it is a constant of type $\tvar{nat}$ represented by the value $0$.
+Expressions of the form $\econst{\psi}$ compile to the denotation associated with $\psi$ if $\psi$ is a constant, as can be seen in Figure 6.
+%%
+%%Operators of higher arity, $n$, are parameterized by a type-level function that takes $n$ denotations (of kind $\kDen$, see below).
+This function must analyze the types associated with these denotations, then return a new denotation composed using the input denotations to give meaning to expressions of the form $\eop{\psi}{\splat{e}{1}{n}}$.
+Note that this implies that operators must have a strictly compositional semantics -- neither expressions nor internal terms can be directly manipulated in the specification language.
+%%
+%%Because type checking and evaluation of function application is not compositional (requiring extending a context and substitution, respectively), it cannot be simulated by an operator definition in $\lamA$ as stated.
+%%
+%%\subsubsection{Denotations}
+%%Although operator implementations take arguments of kind $\kDen$, a denotation itself does not have principal kind $\kDen$.
+Recall that denotations contain a type, and types have kind $\kType{\kappa}$, where $\kappa$ is the kind of the type's index.
+To support the extraction of a type from a denotation, we must ``remember'' its index kind even when it has been packed into a denotation.
+As a result, denotations have principal kind $\kDenk{\kappa}$.
+%%
+%%To allow for operators that are polymorphic over a number of types (but not necessarily {\it all} types), we allow for the index kind to be erased as needed.
+To reintroduce the kind, the $\textsf{idxcase}_{D}$ form is available to perform conditional kind analysis during normalization.
+This can be thought of as a simplified form of subkinding with casts.
+Analagous erasure semantics exist for types themselves, with a corresponding {\sf idxcase} form.
+%%
+%%To ensure that this does not allow for non-terminating terms, we impose a strict positivity restriction on type indices that contain values of kind $\kTypeBlur$.
+This condition, which restricts $\kTypeBlur$ so that it cannot appear to the left of any arrows, is found in many total functional programming languages that feature inductive datatypes, such as Coq \cite{bertot04book} and Agda \cite{norell2007towards}, and we refer the reader to these sources for a detailed discussion.
+%%
+%%Another important reason for the distinction between $\kDen$ and $\kDenk{\kappa}$ is that there must be some mechanism by which an operator specification indicates a type error.
+In $\lamA$, this corresponds to a special value, $\terr$, which is also of kind $\kDen$.
+However, no type is associated with the denotation $\terr$, so it can never be given the kind $\kDenk{\kappa}$.
+The verification and translation rules in Figure 6 ensure that $\terr$ is never propagated past the operator that first produces it.
+%%
+%%\subsection{G\"{o}del's $T$ in $\lamA$}
+%%In Section 2.1, we showed an encoding of natural numbers in $\lamA$, $\psit{nat}$.
+It would be instructive now for the reader to review the examples in that section and derive the various components of the compilation judgement for them.
+For the two negative examples, it is interesting to see why they fail to compile -- the first does not pass kind checking while the second is a user-defined error caught by the definition of $\tvar{s}$.
+%%
+%%We can now complete the encoding of G\"{o}del's $T$ in $\lamA$. The missing piece is the natural recursor. For review, the typing rule for the natural recursor can be written (omitting contexts):
+%%\begin{mathpar}
+%%\inferrule{
+%%	{e_{1}:\tvar{nat}}\\
+%%	{e_{2}:\tau}\\
+%%	{e_{3}:\tvar{nat}\rightarrow\tau\rightarrow{\tau}}
+%%}{
+%%	{\tt natrec}~e_{1}~e_{2}~e_{3} : \tau
+%%}
+%%\end{mathpar}
+%%The corresponding operator definition in $\lamA$, written in a concrete syntax that omits keywords and ``{\sf else err}'' branches is:
+%%\begin{verbatim}
+%%def natrec := forall k. op(fun (d0 d1 d2 : Den) := 
+%%  didxcase d0 of NAT => 
+%%  if d0 = nat => 
+%%  didxcase d1 of k => 
+%%  didxcase d2 of Type[NAT]*(Type[k]*Type[k]) =>
+%%  def t1 := typeof d1 in 
+%%  def t2 := typeof d2 in 
+%%  if fst idx t2 = nat =>
+%%  def s := snd idx t2 in 
+%%  if fst s = snd s => 
+%%  if fst s = t1 => 
+%%  [[(fix f (n :: int) :: rep t1 :=
+%%       if n = 0 then valof(d1) 
+%%       else valof(d2) (n - 1) (f (n - 1))
+%%     ) valof(d0), t1]]) in x
+%%\end{verbatim}
+%%Here, we see that natrec is parameterized by the kind of the index of the type $\tau$ above, corresponding to \texttt{t1} with index kind \texttt{k}.
+The body of the definition amounts to checking that the equational relations between the types for the three arguments that are implied by the inductive definition actually hold.
+The implementation of the natural recursor is a standard recursive definition that mirrors the operational semantics usually given to it.
+If we call this specification $\psit{natrec}$, then the composition of the two specifications produces a specification that is sound and complete against $T$:
+%%$$\psi_{T} := \subst{\psit{natrec}}{\tvar{x}}{\psit{nat}}$$
+%%
+%%One might ask at this point what the benefit is of using this more complex specification style?
+For simple examples like this, it is clearly more complex than necessary.
+However, we will see in Section 4 examples where a purely inductive specification of a language construct requires significantly more machinery than a functional specification.
+In cases where the simpler inductive style is convenient, on the other hand, a sound translation from an inductive specification language to a naive type-level specification would not be difficult to develop.
+Moreover, automatically extracting an efficient implementation from an inductive specification of a semantics is generally far more difficult.
+With type-level specification, we err on the side of greater control whenever safe.
+%%\subsection{Safety of $\lamA$}
+%%As mentioned in the introduction, evaluating complex type-level functions to verify and translate programs is fraught with theoretical peril.
+Both the decidability of type-checking and the metatheoretic and global safety properties of the language itself are at risk of disruption by unwise user specifications.
+Fortunately, the formulation of $\lamA$ allows us to prove two important safety theorems.
+%%
+%%The first guarantees that the compilation judgement of Figure 2, which includes the type checking and translation phases, is decidable.
+%%\begin{theorem}[Decidability of Compilation]
+%%There is a terminating decision procedure that can derive the compilation judgement for any term $\psi$ if it exists, or halt if it does not exist.
+%%\end{theorem}
+%%\noindent
+%%The second guarantees that successful compilation results in a well-typed internal language term that is representationally consistent with its type assignment.
+%%\begin{theorem}[Type-Preserving Compilation]
+%%If 
+%%$$\compiles{\psi}{\mu}{\psiidx}{\kappa}{\psirep}$$ 
+%%then $\mtentails{\cdot}{\mtof{\mu}{\psirep}}$.
+%%\end{theorem}
+%%
+%%\subsubsection{Metatheoretic Properties of the Type-Level Language}
+%%There are three main lemmas that we must show to ensure that compilation is decidable.
+%%\begin{lemma}[Decidability of Kind Checking]
+%%Derivation of the kind assignment judgement $\tentailsXt{\psi}{\kappa}$ is decidable.
+%%\end{lemma}
+%%\begin{lemma}[Decidability of Internal Type Checking]
+%%Derivation of the internal type checking judgement $\mtentailsXt{\mu}{\psi}$ is decidable if $\psi$ is a canonical form for kind $\kIType$.
+%%\end{lemma}
+%%\begin{lemma}[Strong Normalization to Canonical Form] If a closed type-level term $\psi$ is well-kinded, that is 
+%% $\tentails{\cdot}{\cdot}{\cdot}{\tof{\psi}{\kappa}}$,  then $\tevalsX{\psi}{\psi'}{\Sigma}$ for some $\Sigma$ and $\psi'$ is a canonical form of kind $\kappa$.
+%%\end{lemma}
+%%\noindent
+%%We omit statements of the canonical forms lemma and other minor lemmas due to space considerations.
+Lemma 3 is the most difficult to prove but also absolutely critical -- a type-level language that does not guarantee termination will necessarily lead to undecidable type checking.
+Our approach, which we do not have space to detail here, is based on the standard logical relations formulation for strong normalization in System F, with special considerations of type indices.
+The check for strictly positive occurrences of the kind $\kTypeBlur$ ensures that values of kind $\kTypeBlur$ cannot sneak self-references into functions by clever use of {\sf idxcase}, by an argument exactly analagous to those given for total functional programming languages with inductive data types.
+In fact, we'll see in the next section that this analogy between type constructors and inductive data types has significance in practice.
+%%
+%%Using these three lemmas, we can show by simple inductive argument that the verification and compilation rules of Figure 5 are decidable, and thus that Theorem 1 holds.
+%%\subsubsection{Representational Consistency}
+%%Theorem 2 can be shown straightforwardly by noting that the big-step evaluation rules for denotations check representational consistency directly during normalization steps.
+The error context $\Sigma$ tracks whether any denotation constructed during the normalization and verification and translation processes is representationally inconsistent.
+If so, the final evaluation judgement cannot be derived.
+Combined with the representational consistency of the rule for arrow types, described in Section 2.2.2, we have a stronger theorem than the one stated -- that no denotation created during the compilation of the program $\psi$ is representationally inconsistent.
+%%
+%%The issue is that this form of checking amounts to ``dynamic checking'' of this particular invariant at the type-level.
+In particular, we cannot check that a well-kinded specification will produce a valid denotation or $\terr$ for all possible programs that it is applied to -- only those branches of the specification that are exercised for a particular program are checked.
+This means that users will occasionally be faced with specification errors even after the kind checking phase, in addition to any errors arising from problems in their programs.
+%%
+%%The reason we can't statically check specifications for this error is that it is a relational property between the two components of a denotation.
+As such, it requires much more sophisticated support for equational reasoning in the kind system.
+In $\psit{nat}$, for example, proving that even $\tvar{z}$ uses a representationally consistent denotation requires static knowledge that $\trepof{\tvar{nat}}=\tILtint$.
+While this is easily discernable from the specification as given, the simply-kinded system given does not track enough information to prove this kind of property.
+The problem is more clearly evident in the specifications for $\tvar{s}$ and $\tvar{natrec}$, which contain uses of \textsf{valof}.
+To prove that the internal term $1 + \mvalof{\psi}$ is well-typed without evaluating $\psi$ requires a proof that $\trepof{\ttypeof{\psi}}=\tILtint$.
+To prove that the denotation is representationally consistent then requires the same proof that $\trepof{\tvar{nat}}=\tILtint$.
+In Section 3, we give a method for translating type-level terms into Coq, which supports such reasoning.
+In practice, coverage tests to ensure that all branches of a specification are exercised by tests may be sufficient for prototype implementations of language features, with rigorous equational reasoning reserved for those features being deployed more widely or in safety-critical applications.
+Indeed, even in certified compilers like CompCert, the type checking and early translation phases are written in a general-purpose functional language (e.g.
+Ocaml) for practicality \cite{Leroy-Compcert-CACM}.
+%%
+%%\subsubsection{Local vs. Global Safety Guarantees}
+%%While Theorem 2 guarantees that any global properties guaranteed by the internal type system will hold for all compiled programs, this mechanism does not support the introduction of new global safety properties for all possible programs.
+For instance, although the implementation of System $T$ described here can be shown sound and complete, some of the safety properties guaranteed by the constructs of $\psi_{T}$ hold only locally -- in program fragments that only use operators defined in $\psi_{T}$.
+A user could easily introduce a new constant $\tvar{b} := \tconst{\tden{-1}{\tvar{nat}}}$, for example.
+The introduction of this constant invalidates the mapping between the natural numbers of mathematics and the type $\tvar{nat}$, effectively redefining it.
+%%
+%%The purpose of this section was to demonstrate a technique that allows for the concrete specification and modular introduction of new primitives into a language.
+This facilitates experimentation with collections of primitives known (or at least strongly suspected) to be mutually consistent.
+But just as introducing new axioms into a proof assistant can cause problems if not done carefully, composing primitives should also be done with care.
+The hope is that by making experimentation as easy as importing a library and ensuring that it cannot invalidate fundamental properties of the language, different communities will more quickly converge upon powerful collections of mutually consistent primitives.
+%%
+%%In the next section, we sketch a general technique for proving stronger safety and composability theorems about specifications.
+%%\section{Strengthening Specifications with Dependent Kinds}
+%%In the previous section, we gave a minimal calculus that expresses the main ideas of this paper -- a type-level specification language that contains a uniform representation for types and operators.
+Operators were implemented as type-level functions operating on both indexed types and denotations.
+Although the language and compilation process could be shown to be safe in the face of these extensions, the extensions themselves could have errors in them -- we discussed representational consistency, soundness with respect to an independent specification (e.g.
+mathematical natural numbers) and composability of global guarantees.
+%%
+%%These issues all have in common that they require making complex statements about specifications.
+Proof assistants based in type theory such as Coq, Agda, Twelf and others are often used by programming languages researchers for just that task.
+By taking advantage of dependent types married with flexible inductive datatypes, these proof assistants are emerging as powerful tools for certified programming.
+They also possess the fundamental properties required of our metalanguage -- strong normalization.
+%%
+%%To practically realize a language featuring type-level specifications and demonstrate techniques that could be used for certifying these specifications, we propose lifting one of these languages into the type level.
+The key issue is that these languages do not admit a simple embedding for uniform type constructors.
+In most language specifications, types are represented using an inductive data type with a finite set of cases.
+In $\lamA$ however, types can be introduced anywhere.
+This is an instantiation of the well-known expression problem of Wadler \cite{wadler1998expression}.
+%%
+%%A number of techniques have been proposed in the literature to resolve this issue in general-purpose functional languages.
+Open data types are one popular approach \cite{conf/ppdp/LohH06}.
+The indexed types in $\lamA$ can be considered an open data kind where each possible index kind generates a new constructor.
+Each constructor must contain a slot to place the associated internal type's representation.
+We use this observation to construct a translation from a language in the style of $\lamA$ to Gallina, the dependently-typed language of the Coq proof assistant.
+The kinds that are possible for indices can be read statically from a type-level term.
+Index case statements containing closed kinds can thus be translated into \texttt{match} statements containing two cases: one for to the corresponding constructor, and one matching all other constructors.
+Index case statements containing kind variables (e.g.
+natrec) require a more sophisticated approach, however.
+The case kind must be unified against all possible closed index kinds and a match case must be generated for each one where unification succeeds.
+%%
+%%All the remaining constructs needed by $\lamA$ can be expressed using standard inductive data types.
+By also implementing the language and compiler itself in Gallina and providing a specification of the semantics of the internal language, type-level terms can be used to reason about the behavior of operator implementations in nearly arbitrary detail.
+A proof-passing approach to representational consistency as discussed in Section 2.4.3  can guarantee that users will only ever see errors originating in their own expressions because specifications themselves have been modularly verified.
+%%
+%%To prove strong composability results about a collection of types and primitives, there are a number of possible approaches.
+For the primitives of $\psi_{T}$, for example, it can be seen that it does not interfere with other primitives -- all denotations are assigned either a newly introduced type -- $\tvar{nat}$ -- or an external type -- ${\tt t1}$ -- paired with an internal term taken directly from an input denotation to the operator and also having type {\tt t1}, which we can inductively assume is semantically consistent.
+Proving such a theorem, while not trivial, is well within the capabilities of Coq and other proof assistants.
+Our choice of Coq is largely motivated by familiarity as well as Coq's sophisticated program extraction capabilities.
+%%
+%%\section{Examples}
+%%We now give several examples of more sophisticated language constructs and how they may be implemented using this approach.
+We assume a type-level specification language structured like $\lamA$ but enriched with general inductive data kinds, though not necessarily dependent kinds.
+We assume kind inference when it is known to be decidable in System F (e.g.
+index kind declarations are omitted.) We also assume standard structures common to typed intermediate languages are available in the internal language.
+As in Section 2, missing else clauses are assumed to return \texttt{err}.
+%%
+%%\subsection{Record and Object Types}
+%%A record type can be characterized by a collection of static field names mapped to other types.
+They are commonly implemented using some flat, heterogenous, untagged data structure, because the offsets can be computed statically.
+We can therefore declare the type family of records as a type-level function that produces a type indexed by this specification and represented by a generic flat heterogeneous structure.
+%%
+%%\begin{verbatim}
+%%fun (spec : list[(str, Type)]) := 
+%%  type[RECORD spec] :: flat[map spec 
+%%    fun (s:str, t:Type) => (rep t)]
+%%\end{verbatim}
+%%
+%%To extract a value from a record, we need an appropriate operator.
+We can implement this as a type-level function that produces an operator when given an index string.
+We assume an isRec function has been defined to check that the representation type is consistent with the index as defined above.
+We also assume that type-level numbers can be translated to internal language numbers using the \verb|#| operator.
+Several other helper functions are also assumed where the meaning is apparent.
+%%
+%%\begin{verbatim}
+%%fun get(idx : str) := op(fun d : Den.
+%%  didxcase d of RECORD =>
+%%  if isRec typeof d =>
+%%  def spec = open idx typeof d
+%%  if containsIdx spec idx =>  
+%%    [[valof(d)[#(getOffset spec idx)], 
+%%      typeForIdx spec idx]])
+%%\end{verbatim}
+%%
+%%An analagous set function can be implemented.
+Object types are like records, except they have additional metadata expressing inheritance relationships and other annotations.
+This additional information could also be packaged into the type index in a straightforward way, allowing for languages supporting a variety of different object protocols.
+%%
+%%\subsection{Scientific Units}
+%%An example of a more specialized application domain is scientific unit checking.
+Numerical values (and sometimes collections of numerical values) can be annotated with dimensions (e.g.
+mass) and specific units of measure (e.g.
+grams).
+Addition and subtraction must operate on values with identical dimension and respect conversion rules if operating on heterogeneous units (e.g.
+grams + kilograms).
+Multiplication and division, on the other hand, cause the formation and elimination of compound units (e.g.
+meters/sec.).
+Although standalone type systems have been developed which respect these rules (e.g.
+\cite{conf/cefp/Kennedy09}), they are quite difficult to implement in terms of conventional language features, particularly if users hope to erase these annotations at run-time for performance reasons.
+Few languages have enough scientific users to justify adding this feature to the core language however.
+%%
+%%We can represent compound scientific types as being indexed by type-level lists of multiplicand and divisor units, where units are type-level values associated with dimensions.
+The corresponding rules can then be encoded directly as operators.
+%%
+%%\begin{verbatim}
+%%def unitNum = fun (mult div : list[SIUnit]) :=
+%%  type[SINUM mult, div] :: int
+%%\end{verbatim}
+%%\noindent
+%%We omit the details of the operator implementations.
+%%
+%%\subsection{Length-Indexed Lists}
+%%Length-indexed lists are a common example for dependently-typed languages.
+With type-level numbers, such lists are relatively straightforward to encode.
+All operators use standard arithmetic operations on the index to ensure that the length is accurately maintained.
+%%
+%%\begin{verbatim}
+%%def list = fun (length : nat, t : Type) := 
+%%  type[LIST length, t] :: array[rep t]
+%%\end{verbatim}
+%%
+%%Other common examples of dependently-typed data structures can be encoded similarly, though the dependency in this case is only on type-level values (or perhaps denotations), but not run-time values.
+%%
+%%\subsubsection{Interoperability Layers}
+%%A type-safe interoperability layer can be derived from an unsafe interoperability layer encoded by implementing the type system of the target language, with type representations corresponding to the core data structures available in the unsafe interoperability layer.
+%%\section{Related Work}
+%%\subsection{Type-Level Computation} %Haskell, Ur and $\Omega$mega
+%%System $\text{F}_{\omega}$ with simple case analysis provides the basis of type-level computation in Haskell (where type-level functions are called type families \cite{Chakravarty:2005:ATC}).
+Ur uses type-level records and names to support typesafe metaprogramming, with applications to web programming \cite{conf/pldi/Chlipala10}.
+$\Omega$mega adds algebraic data types at the type-level, using these to increase the expressive power of algebraic data types at the expression level \cite{conf/cefp/SheardL07}.
+Dependently-typed languages blur the traditional phase separation between types and expressions, so type-level computation is often implicitly used (though not always in its most general form, e.g.
+Deputy \cite{conf/icfp/ChenX05}, ATS \cite{conf/esop/ConditHAGN07}.)
+%%
+%%\subsection{Run-Time Indirection}
+%%{\it Operator overloading} \cite{vanWijngaarden:Mailloux:Peck:Koster:Sintzoff:Lindsey:Meertens:Fisker:acta:1975} and {\it metaobject dispatch} \cite{Kiczales91} are run-time protocols that translate operator invocations into function calls.
+The function is typically selected according to the type or value of one or more operands.
+These protocols share the notion of {\it inversion of control} with type-level specification.
+However, type-level specification is a {\it compile-time} protocol focused on enabling specialized verification and implementation strategies, rather than simply enabling run-time indirection.
+%%
+%%\subsection{Term Rewriting Systems}
+%%Many languages and tools allow developers to rewrite expressions according to custom rules.
+These can broadly be classified as {\it term rewriting systems}.
+Macro systems, such as those characteristic of the LISP family of languages \cite{mccarthy1978history}, are the most prominent example.
+Some compile-time metaprogramming systems also allow users to manipulate syntax trees (e.g.
+MetaML \cite{Sheard:1999:UMS}), and external rewrite systems also exist for many languages.
+%%These facilities differ from type-level specification in one or more of the following ways:
+%%
+%%\begin{enumerate}
+%%\item In type-level specification, the type of a value is determined separately from its representation; in fact, the same representation may be generated by multiple types. 
+%%\item We draw a distinction between the metalanguage, used to specify types and compile-time logic, the source grammar, used to describe run-time behavior, and the internal language, used to implement this behavior.
+Term rewriting systems generally do not draw this distinction.
+By doing so, each component language can be structured and constrained as appropriate for its distinct role, as we show.
+%%%\item With type-level specification, dispatch to a type-level function occurs implicitly on the basis of the structure of an expression.
+In contrast, most term-rewriting systems operate by  explicit invocation of a macro or specialized syntax.
+Some LISP macro systems have explored pattern-based dispatch (e.g.
+A*\cite{fowler2010domain}, EPP\cite{fowler2010domain}) and macro systems for object-oriented languages, like OpenC++ \cite{fowler2010domain} and OpenJava \cite{fowler2010domain}, do offer a somewhat limited form of operation-based dispatch.
+%%\item Many common macro systems and metaprogramming facilities operate at run-time.
+Compilers for some forms of LISP employ aggressive compile-time specialization techniques to attempt to minimize this overhead.
+Static and staged term-rewriting systems also exist (e.g.
+OpenJava\cite{TatM:OpenJCBMSJ}, Template Haskell\cite{SheardPeytonJones:Haskell-02}, MetaML \cite{Sheard:1999:UMS} and others).
+%%\end{enumerate}
+%%
+%%\subsection{Language Frameworks}
+%%When the mechanisms available in an existing language prove insufficient, researchers and domain experts must design a new language.
+A number of tools have been developed to assist with this task, including compiler generators, language workbenches and domain-specific language frameworks (cf \cite{fowler2010domain}).
+%%
+%%A major barrier to adoption is the fact that interoperability is intrinsically problematic.
+Even languages which target a common platform, such as the Java Virtual Machine, can only interact using its limited set of primitives.
+Specialized typing rules are not checked at language boundaries, performance often suffers, and the syntax can be unnatural, particularly for languages which differ significantly from the platform's native language (e.g.
+Java).
+%%
+%%Instead of focusing on defining standalone languages, type-level specification gives greater responsibility in a granular manner to libraries.
 In this way, a range of constructs can coexist within the same program and, assuming that it can be shown by some method that various constructs are safely composable, be mixed and matched.
@@
 Nevertheless, as $\lamA$ shows, a simple grammar can be used quite flexibly.
-%\subsection{Extensible Compilers}
-%An alternative methodology is to implement language features granularly as compiler extensions.
+%%\subsection{Extensible Compilers}
+%%An alternative methodology is to implement language features granularly as compiler extensions.
 As discussed in Section 1, existing designs suffer from the same problems related to composability, modularity\-, safety and security as extensible languages, while also adding the issue of language fragmentation.
-%
-%Type-level specification can in fact be implemented within a compiler, rather than provided as a core language feature.
+%%
+%%Type-level specification can in fact be implemented within a compiler, rather than provided as a core language feature.
 This would resolve some of the issues, as described in this paper.
@@
 We also avoid the fragmentation issue.
-%\subsection{Combining Logics and Specification Languages}
-%Programming language specifications can be understood as formal logics.
+%%\subsection{Combining Logics and Specification Languages}
+%%Programming language specifications can be understood as formal logics.
 The question of {\it combining logics} \cite{sep-logic-combining} can be formulated as follows: given two logics, $\mathcal{L}_{1}$ and $\mathcal{L}_{2}$, how can we produce a minimal new logic, $\mathcal{L}$, such that it includes both $\mathcal{L}_{1}$ and $\mathcal{L}_{2}$?
-%
-%The major approaches are based in category theory and include {\it fibring} \cite{acs:css:ccal:98a} and the framework of {\it institutions} \cite{Diaconescu-Futatsugi01}.
+%%
+%%The major approaches are based in category theory and include {\it fibring} \cite{acs:css:ccal:98a} and the framework of {\it institutions} \cite{Diaconescu-Futatsugi01}.
 Many variants of these approaches exist, addressing different formulation styles and focusing on guaranteeing different properties about the combined logics by introducing various constraints.
-%
-%Several {\it specification languages} (or {\it logical frameworks}) based on these theoretical formulations exist, including the OBJ family of languages (e.g.
+%%
+%%Several {\it specification languages} (or {\it logical frameworks}) based on these theoretical formulations exist, including the OBJ family of languages (e.g.
 CafeOBJ \cite{Diaconescu-Futatsugi01}).
@@
 The  language itself specifies which verification and execution strategies are used.
-%
-%Type-determined compilation takes a more concrete approach to the problem, focusing on combining {\it implementations} of different\- logics, rather than simply their specifications.
+%%
+%%Type-determined compilation takes a more concrete approach to the problem, focusing on combining {\it implementations} of different\- logics, rather than simply their specifications.
 In other words, it focuses on combining {\it type checkers} and {\it implementation strategies} rather than more abstract representations of a language's type system and dynamic semantics.
 In Section 4, we outlined a preliminary approach based on proof assistant available for the type-level language to unify these approaches, and we hope to continue this line of research in future work.
-%
-%
-%\section{Future Work}
-%\subsection{General-Purpose Language Development}
-%We have started to explore applications of this work to scientific programming, building a variant of this system (with a weaker metalanguage than described in Section 4) that targets the OpenCL programming language for GPU computing.
+%%
+%%
+%%\section{Future Work}
+%%\subsection{General-Purpose Language Development}
+%%We have started to explore applications of this work to scientific programming, building a variant of this system (with a weaker metalanguage than described in Section 4) that targets the OpenCL programming language for GPU computing.
 The entirety of the C99 can be implemented using a variant of this approach that is not strictly compositional (an approach that can lead to non-termination).
 We will continue this work while also continuing to develop the more theoretically sound foundations described here.
-%
-%Although we have implemented several powerful data structures, more thorough  case studies must be performed with such a language before we can definitively say that this approach could lead to a practical programming language.
+%%
+%%Although we have implemented several powerful data structures, more thorough  case studies must be performed with such a language before we can definitively say that this approach could lead to a practical programming language.
 We conjecture that this process will motivate further research questions related to this protocol as well.
-%
-%\subsection{Integration with Tools}
-%Developers increasingly turn to integrated development environments, debuggers and refactoring tools to assist with their programming tasks.
+%%
+%%\subsection{Integration with Tools}
+%%Developers increasingly turn to integrated development environments, debuggers and refactoring tools to assist with their programming tasks.
 Many of these tools are built against a fixed language definition.
 There remain significant challenges in understanding how to integrate language designs based on type-level specification into these tools effectively.
-%
-%\section{Conclusion}
-%In conclusion, we have shown a new new approach to language extensibility that works at the type-level.
+%%
+%%\section{Conclusion}
+%%In conclusion, we have shown a new new approach to language extensibility that works at the type-level.
 This approach is flexible but can also shown to be safe, in that it respects global safety properties of the language and does not negative impact the integrity of compilation.
 We gave a simple core calculus for this approach that we hope will be useful for distilling the essential concept, sketched an extension of the calculus that can be used to prove stronger metatheoretic guarantees and sketched several examples of relatively complex language constructs that admit implementation using this approach.
-%
-%\acks
-%%The author is grateful to Jonathan Aldrich, Robert Bocchino and anonymous referees for their useful suggestions.
+%%
+%%\acks
+%%%The author is grateful to Jonathan Aldrich, Robert Bocchino and anonymous referees for their useful suggestions.
 This work was funded by the DOE Computational Science Graduate Fellowship under grant number DE-FG02-97ER25308.
-%%Acknowledgments, if needed.
-%
-%% We recommend abbrvnat bibliography style.
-%
-%\bibliographystyle{abbrvnat}
-%
-%% The bibliography should be embedded for final submission.
-%
-%\bibliography{../research}
-%%\softraggedright
-%%P. Q. Smith, and X. Y. Jones. ...reference text...
-%
-%%\appendix
-%%The grammar of $\lamTDC$ is:
-%%\begin{eqnarray*}
-%%\kappa & ::= & \kinds\\
-%%\tau & ::= & \metaterms\\
-%%e & ::= & \sourcegrammar\\
-%%\tvarCtx & ::= & \datatypes\\
-%%\evarCtx & ::= & \ilgrammar
-%%\end{eqnarray*}
-%%Note that $\intlit$ is a placeholder for any integer literal and $x$ is a placeholder for variable names. In this paper, we use strings of italicized, lowercase letters (e.g. {\it gram}).
-%%\subsection{Intermediate Language Statics}
-%%Intermediate language terms, $\evarCtx$, form a simple expression language.
+%%%Acknowledgments, if needed.
+%%
+%%% We recommend abbrvnat bibliography style.
+%%
+%%\bibliographystyle{abbrvnat}
+%%
+%%% The bibliography should be embedded for final submission.
+%%
+%%\bibliography{../research}
+%%%\softraggedright
+%%%P. Q. Smith, and X. Y. Jones. ...reference text...
+%%
+%%%\appendix
+%%%The grammar of $\lamTDC$ is:
+%%%\begin{eqnarray*}
+%%%\kappa & ::= & \kinds\\
+%%%\tau & ::= & \metaterms\\
+%%%e & ::= & \sourcegrammar\\
+%%%\tvarCtx & ::= & \datatypes\\
+%%%\evarCtx & ::= & \ilgrammar
+%%%\end{eqnarray*}
+%%%Note that $\intlit$ is a placeholder for any integer literal and $x$ is a placeholder for variable names. In this paper, we use strings of italicized, lowercase letters (e.g. {\it gram}).
+%%%\subsection{Intermediate Language Statics}
+%%%Intermediate language terms, $\evarCtx$, form a simple expression language.
 The intermediate language's types, $\tvarCtx$, are called {\it data types}.
 The statics take the form $\evarCtx :: \tvarCtx$.
-%%\begin{mathpar}
-%%\intlitT
-%%~~~~
-%%\addT
-%%\end{mathpar}\begin{mathpar}
-%%\tupleT\end{mathpar}\begin{mathpar}
-%%\fstT
-%%~~~~
-%%\sndT
-%%~~~~
-%%\faultT
-%%\end{mathpar}
-%%\subsection{Metalanguage Statics}
-%%Type-level terms, $\tau$, form a simply typed lambda calculus. The types of the metalanguage, $\kappa$, are called {\it kinds}. 
-%%Statics take the form $\entails{\ctx}{\tau \sim \kappa}$, where the context, $\ctx$, operates as a set of assignments of kinds to variables and $\ctx \uplus x \sim \tau$ replaces any previous assignment for $x$ in $\ctx$ with $x\sim\tau$.
-%%\paragraph{Standard Terms}
-%%\newcommand{\axiom}[1]{\inferrule{ }{#1}}
-%%\begin{mathpar}
-%%\varT~~~~
-%%\kfnIT\\
-%%\kfnET
-%%\end{mathpar}
-%%For convenience, a simple ${\sf let}$ form is included.
-%%\begin{mathpar}
-%%\letT
-%%\end{mathpar}
-%%The ${\sf if}$ form checks whether two terms are $\beta$-equivalent. The two branches must evaluate to terms of the same kind.
-%%\begin{mathpar}
-%%\ifT
-%%\end{mathpar}
-%%The remaining statics are described in Section 3. We list them again, and give each a name, below.
-%%\paragraph{Type Construction}
-%%\begin{mathpar}
-%%\typeT\\
-%%\errT
-%%\end{mathpar}
-%%\paragraph{Code Generation Primitives}
-%%\begin{mathpar}
-%%\ilrepT\\
-%%\iladdT\\
-%%\iltupleT\\
-%%\ilfstT~~~~
-%%\ilsndT
-%%\end{mathpar}
-%%
-%%\paragraph{Source Parsing Rules}
-%%\begin{mathpar}
-%%\numtreePT~~~~\newtreePT\\
-%%\optreePT
-%%\end{mathpar}
-%%\paragraph{Syntax Tree Representations}
-%%\begin{mathpar}
-%%\numtreeT\\\newtreeT\end{mathpar}{\begin{mathpar}\optreeT\\\visitT
-%%\end{mathpar}
-%%\paragraph{Source Program Representation}
-%%\begin{mathpar}
-%%\sourcetT
-%%\end{mathpar}
-%%\subsection{Metalanguage Dynamics}
-%%We now list the dynamics of the metalanguage.
-%%\paragraph{Standard Terms}
-%%\begin{mathpar}
-%%\inferrule{ }{
-%%	\final{\kfn{x}{\kappa}{\tau}}
-%%}
-%%~~~~
-%%\inferrule{
-%%	\tau_0 \reduces \tau_0'
-%%}{
-%%	\tau_0~\tau_1 \reduces \tau_0'~\tau_1
-%%}
-%%\\
-%%\inferrule{ }{
-%%	(\kfn{x}{\kappa}{\tau})~\tau_0 \reduces \subst{\tau_0}{x}{\tau}
-%%}
-%%~~~~
-%%\inferrule{ }{
-%%	\klet{x}{\tau_0}{\tau_1} \reduces \subst{\tau_0}{x}{\tau_1}
-%%}\\
-%%\inferrule{\tau_0 \reduces \tau_0'}{
-%%	\ifeq{\tau_0=\tau_1}{\tau_\true}{\tau_\false} \reduces \ifeq{\tau_0' = \tau_1}{\tau_\true}{\tau_\false}
-%%}\\
-%%\inferrule{\final{\tau_0}\\\tau_1 \reduces \tau_1'}{
-%%	\ifeq{\tau_0=\tau_1}{\tau_\true}{\tau_\false} \reduces \ifeq{\tau_0 = \tau_1'}{\tau_\true}{\tau_\false}
-%%}\\
-%%\inferrule{\final{\tau_0}\\\final{\tau_1}\\\tau_0 = \tau_1}{
-%%	\ifeq{\tau_0=\tau_1}{\tau_\true}{\tau_\false} \reduces \tau_\true
-%%}\\
-%%\inferrule{\final{\tau_0}\\\final{\tau_1}\\\tau_0 \neq \tau_1}{
-%%	\ifeq{\tau_0=\tau_1}{\tau_\true}{\tau_\false} \reduces \tau_\false
-%%}
-%%\end{mathpar}
-%%$\tau_0 = \tau_1$ and $\tau_0 \neq \tau_1$ denote syntactic equality and inequality.
-%%\paragraph{Type Construction}
-%%\begin{mathpar}
-%%\inferrule{ }{
-%%	\final{\typedefstd}
-%%}
-%%~~~~
-%%\inferrule{ }{
-%%	\final{\err}
-%%}
-%%\end{mathpar}
-%%\paragraph{Code Generation Primitives}
-%%\begin{mathpar}
-%%\inferrule{ }{
-%%	\final{\evarCtx[\evarCtx]}
-%%}
-%%~~~~
-%%\inferrule{
-%%	\tau_L \reduces \tau_L'
-%%}{
-%%	\evarCtx_{+}(\tau_L, \tau_R) \reduces \evarCtx_{+}(\tau_L', \tau_R)
-%%}\end{mathpar}\begin{mathpar}
-%%\inferrule{
-%%	\tau_R \reduces \tau_R'	
-%%}{
-%%	\evarCtx_{+}(\evarCtx[\evarCtx_L], \tau_R) \reduces \evarCtx_{+}(\evarCtx[\evarCtx_L], \tau_R')
-%%}\end{mathpar}\begin{mathpar}
-%%\inferrule{ }{
-%%	\evarCtx_{+}(\evarCtx[\evarCtx_L], \evarCtx[\evarCtx_R]) \reduces \evarCtx[\evarCtx_L + \evarCtx_R]
-%%}\end{mathpar}\begin{mathpar}
-%%\inferrule{
-%%	\tau_L \reduces \tau_L'
-%%}{
-%%	\evarCtx_{\times}(\tau_L, \tau_R) \reduces \evarCtx_{\times}(\tau_L', \tau_R)
-%%}\end{mathpar}\begin{mathpar}
-%%\inferrule{
-%%	\tau_R \reduces \tau_R'	
-%%}{
-%%	\evarCtx_{\times}(\evarCtx[\evarCtx_L], \tau_R) \reduces \evarCtx_{\times}(\evarCtx[\evarCtx_L], \tau_R')
-%%}\end{mathpar}\begin{mathpar}
-%%\inferrule{ }{
-%%	\evarCtx_{\times}(\evarCtx[\evarCtx_L], \evarCtx[\evarCtx_R]) \reduces \evarCtx[(\evarCtx_L, \evarCtx_R)]
-%%}\end{mathpar}\begin{mathpar}
-%%\inferrule{\tau_\times \reduces \tau_\times'}{
-%%	\evarCtxfst{\tau_\times} \reduces \evarCtxfst{\tau_\times'}
-%%}
-%%~~~~
-%%\inferrule{ }{
-%%	\evarCtxfst{\evarCtx[(\evarCtx_L, \evarCtx_R)]} \reduces \evarCtx[\evarCtx_L]
-%%}\end{mathpar}\begin{mathpar}
-%%\inferrule{\tau_\times \reduces \tau_\times'}{
-%%	\evarCtxsnd{\tau_\times} \reduces \evarCtxsnd{\tau_\times'}
-%%}
-%%~~~~
-%%\inferrule{ }{
-%%	\evarCtxsnd{\evarCtx[(\evarCtx_L, \evarCtx_R)]} \reduces \evarCtx[\evarCtx_R]
-%%}
-%%\end{mathpar}
-%%\paragraph{Syntax Tree Representations}
-%%\begin{mathpar}
-%%\inferrule{
-%%	\tau_n \reduces \tau_n'
-%%}{
-%%	\numtree{\tau_n} \reduces \numtree{\tau_n'}
-%%}~~~~
-%%\axiom{\final{\numtree{\evarCtx[\intlit]}}}\end{mathpar}\begin{mathpar}
-%%\inferrule{
-%%	\tau_\star \reduces \tau_\star'
-%%}{
-%%	\newtreestd \reduces \newtree{\tau_\star'}{\tau_\pi}
-%%}~~~~
-%%\inferrule{
-%%	\final{\tau_\star}\\
-%%	\tau_\pi \reduces \tau_\pi'
-%%}{
-%%	\newtreestd \reduces \newtree{\tau_\star}{\tau_\pi'}
-%%}\end{mathpar}\begin{mathpar}
-%%\inferrule{
-%%	\final{\tau_\star}\\
-%%	\final{\tau_\pi}
-%%}{
-%%	\final{\newtreestd}
-%%}\end{mathpar}\begin{mathpar}
-%%\inferrule{
-%%	\tauL \reduces \tauL'
-%%}{
-%%	\optreestd \reduces \optree{\tauL'}{\tauR}
-%%}~~~~
-%%\inferrule{
-%%	\final{\tauL}\\
-%%	\tauR \reduces \tauR'
-%%}{
-%%	\optreestd \reduces \optree{\tauL}{\tauR'}
-%%}\end{mathpar}\begin{mathpar}
-%%\inferrule{
-%%	\final{\tauL}\\
-%%	\final{\tauR}
-%%}{
-%%	\final{\optreestd}
-%%}
-%%\end{mathpar}
-%%\begin{mathpar}
-%%\inferrule{
-%%	\taupi \reduces \taupi'
-%%}{
-%%	\kvisitstd \reduces \kvisit{\taupi'}{\tau_{\#}}{\tauadd}{\taunew}
-%%}\end{mathpar}\begin{mathpar}
-%%\axiom{\kvisit{\numtreestd}{\tau_{\#}}{\tauadd}{\taunew} \reduces \tau_{\#}~\tau_n}\end{mathpar}\begin{mathpar}
-%%\axiom{\kvisit{\newtreestd}{\tau_{\#}}{\tauadd}{\taunew} \reduces \taunew~\tau_\star~\taupi}\end{mathpar}\begin{mathpar}
-%%\axiom{\kvisit{\optreestd}{\tau_{\#}}{\tauadd}{\taunew} \reduces \tauadd~\tauL~\tauR}\end{mathpar}
-%%\paragraph{Source Program Representations}
-%%\begin{mathpar}
-%%\axiom{\final{\source{\intlit}}}\\
-%%\inferrule{
-%%	\tau \reduces \tau'
-%%}{
-%%	\source{\newobj{\tau}{e}} \reduces \source{\newobj{\tau'}{e}}
-%%}\\
-%%\inferrule{
-%%	\final{\tau}\\
-%%	\source{e} \reduces \source{e'}
-%%}{
-%%	\source{\newobj{\tau}{e}} \reduces \source{\newobj{\tau}{e'}}
-%%}\\
-%%\inferrule{
-%%	\final{\tau}\\
-%%	\final{\source{e}}
-%%}{
-%%	\final{\source{\newobj{\tau}{e}}}
-%%}\\
-%%\inferrule{
-%%	\source{e_L} \reduces \source{e_L'}
-%%}{
-%%	\source{\absop{e_L}{e_R}} \reduces \source{\absop{e_L'}{e_R}}
-%%}\\
-%%\inferrule{
-%%	\final{\source{e_L}}\\
-%%	\source{e_R} \reduces \source{e_R'}
-%%}{
-%%	\source{\absop{e_L}{e_R}} \reduces \source{\absop{e_L}{e_R'}}
-%%}\\
-%%\inferrule{
-%%	\final{\source{e_L}}\\
-%%	\final{\source{e_R}}
-%%}{
-%%	\final{\source{\absop{e_L}{e_R}}}
-%%}
-%%\end{mathpar}
-%%\paragraph{Substitution Dynamics}
-%%\subparagraph{Standard Terms}
-%%\begin{mathpar}
-%%\inferrule{ }{
-%%	\subst{\tau}{x}{x} \reduces \tau
-%%}
-%%~~~~
-%%\inferrule{x \neq y}{
-%%	\subst{\tau}{x}{y} \reduces y
-%%}\end{mathpar}\begin{mathpar}
-%%\inferrule{ }{
-%%	\subst{\tau}{x}{\kfn{x}{\kappa}{\tau_0}} \reduces \kfn{x}{\kappa}{\tau_0}
-%%}\end{mathpar}\begin{mathpar}
-%%\inferrule{
-%%	x \neq y\\
-%%	\subst{\tau}{x}{\tau_0} \reduces \tau_0'	
-%%}{
-%%	\subst{\tau}{x}{\kfn{y}{\kappa}{\tau_0}} \reduces \kfn{y}{\kappa}{\tau_0'}
-%%}\end{mathpar}\begin{mathpar}
-%%\inferrule{
-%%	\subst{\tau}{x}{\tau_0} \reduces \tau_0'\\
-%%	\subst{\tau}{x}{\tau_1} \reduces \tau_1'
-%%}{
-%%	\subst{\tau}{x}{\tau_0~\tau_1} \reduces \tau_0'~\tau_1'
-%%}\end{mathpar}\begin{mathpar}
-%%\inferrule{
-%%	\subst{\tau}{x}{\tau_0} \reduces \tau_0'\\
-%%}{
-%%	\subst{\tau}{x}{\klet{x}{\tau_0}{\tau_1}} \reduces \klet{x}{\tau_0'}{\tau_1}
-%%}\end{mathpar}\begin{mathpar}
-%%\inferrule{
-%%	x \neq y\\
-%%	\subst{\tau}{x}{\tau_0} \reduces \tau_0'\\
-%%	\subst{\tau}{x}{\tau_1} \reduces \tau_1'
-%%}{
-%%	\subst{\tau}{x}{\klet{y}{\tau_0}{\tau_1}} \reduces \klet{y}{\tau_0'}{\tau_1'}
-%%}\end{mathpar}\begin{mathpar}
-%%\inferrule{
-%%	\subst{\tau}{x}{\tau_0} \reduces \tau_0'~~~~
-%%	\subst{\tau}{x}{\tau_1} \reduces \tau_1'~~~~
-%%	\subst{\tau}{x}{\tau_\true} \reduces \tau_\true'~~~~
-%%	\subst{\tau}{x}{\tau_\false} \reduces \tau_\false'
-%%}{
-%%	\subst{\tau}{x}{\ifeq{\tau_0 = \tau_1}{\tau_\true}{\tau_\false}} \reduces \ifeq{\tau_0' = \tau_1'}{\tau_\true'}{\tau_\false'}
-%%}\end{mathpar}
-%%\subparagraph{Type Construction}
-%%\begin{mathpar}
-%%\inferrule{
-%%	\subst{\tau}{x}{\taunewg} \reduces \hat{\tau}_{\sf new}^\evarCtx\\
-%%	\subst{\tau}{x}{\tauaddg} \reduces \hat{\tau}_\diamond^\star\\
-%%	\subst{\tau}{x}{\tauaddt} \reduces \hat{\tau}_\diamond^\evarCtx
-%%}{
-%%	\subst{\tau}{x}{\typedefstd} \reduces \subst{\tau}{x}{\typedef{a}{\hat{\tau}_{\sf new}^\evarCtx}{\hat{\tau}_\diamond^\star}{\hat{\tau}_\diamond^\evarCtx}{\tvarCtxmap}{f}{g}}
-%%}\end{mathpar}\begin{mathpar}
-%%\inferrule{ }{
-%%	\subst{\tau}{x}{\err} \reduces \err
-%%}\end{mathpar}
-%%\subparagraph{Code Generation Primitives}
-%%\begin{mathpar}
-%%\inferrule{ }{
-%%	\subst{\tau}{x}{\evarCtx[\evarCtx]} \reduces \evarCtx[\evarCtx]
-%%}\end{mathpar}\begin{mathpar}
-%%\inferrule{
-%%	\subst{\tau}{x}{\tau_L} \reduces \tau_L'\\
-%%	\subst{\tau}{x}{\tau_R} \reduces \tau_R'
-%%}{
-%%	\subst{\tau}{x}{\evarCtx_{+}(\tau_L, \tau_R)} \reduces \evarCtx_{+}(\tau_L', \tau_R')
-%%}\end{mathpar}\begin{mathpar}
-%%\inferrule{
-%%	\subst{\tau}{x}{\tau_L} \reduces \tau_L'\\
-%%	\subst{\tau}{x}{\tau_R} \reduces \tau_R'
-%%}{
-%%	\subst{\tau}{x}{\evarCtx_{\times}(\tau_L, \tau_R)} \reduces \evarCtx_{\times}(\tau_L', \tau_R')
-%%}\end{mathpar}\begin{mathpar}
-%%\inferrule{
-%%	\subst{\tau}{x}{\tau_\times} \reduces \tau_\times'
-%%}{
-%%	\subst{\tau}{x}{\evarCtxfst{\tau_\times}} \reduces \evarCtxfst{\tau_\times'}
-%%}
-%%~~~~
-%%\inferrule{
-%%	\subst{\tau}{x}{\tau_\times} \reduces \tau_\times'
-%%}{
-%%	\subst{\tau}{x}{\evarCtxsnd{\tau_\times}} \reduces \evarCtxsnd{\tau_\times'}
-%%}\end{mathpar}
-%%\subparagraph{Syntax Tree Representations}
-%%\begin{mathpar}
-%%\inferrule{
-%%	\subst{\tau}{x}{\tau_n} \reduces \tau_n'
-%%}{
-%%	\subst{\tau}{x}{\numtreestd} \reduces \numtree{\tau_n'}
-%%}
-%%~~~~
-%%\inferrule{
-%%	\subst{\tau}{x}{\tau_\star} \reduces \tau_\star'\\
-%%	\subst{\tau}{x}{\tau_\pi} \reduces \tau_\pi'
-%%}{
-%%	\subst{\tau}{x}{\newtreestd} \reduces \newtree{\tau_\star'}{\tau_\pi'}
-%%}\end{mathpar}\begin{mathpar}
-%%\inferrule{
-%%	\subst{\tau}{x}{\tau_L} \reduces \tau_L'\\
-%%	\subst{\tau}{x}{\tau_R} \reduces \tau_R'
-%%}{
-%%	\subst{\tau}{x}{\optreestd} \reduces \optree{\tau_L'}{\tau_R'}
-%%}\end{mathpar}\begin{mathpar}
-%%\inferrule{
-%%	\subst{\tau}{x}{\tau_\pi} \reduces \tau_\pi'~~~~
-%%	\subst{\tau}{x}{\tau_{\#}} \reduces \tau_{\#}'\\
-%%	\subst{\tau}{x}{\tau_{\sf new}} \reduces \tau_{\sf new}'~~~~
-%%	\subst{\tau}{x}{\tau_\diamond} \reduces \tau_\diamond
-%%}{
-%%	\subst{\tau}{x}{\kvisit{\tau_\pi}{\tau_{\#}}{\tau_\diamond}{\tau_{\sf new}}} \reduces
-%%	\kvisit{\tau_\pi'}{\tau_{\#}'}{\tau_\diamond'}{\tau_{\sf new}'}
-%%}\end{mathpar}
-%%\subparagraph{Source Program Representations}
-%%\begin{mathpar}
-%%\inferrule{ }{
-%%	\subst{\tau}{x}{\source{\intlit}} \reduces \source{\intlit}
-%%}\end{mathpar}\begin{mathpar}
-%%\inferrule{
-%%	\subst{\tau}{x}{\tau_\star} \reduces \tau_\star'\\
-%%	\subst{\tau}{x}{\source{e}} \reduces \source{e'}
-%%}{
-%%	\subst{\tau}{x}{\source{\newobj{\tau_\star}{e}}} \reduces \source{\newobj{\tau_\star'}{e'}}
-%%}\end{mathpar}\begin{mathpar}
-%%\inferrule{
-%%	\subst{\tau}{x}{\source{e_L}} \reduces \source{e_L'}\\
-%%	\subst{\tau}{x}{\source{e_R}} \reduces \source{e_R'}
-%%}{
-%%	\subst{\tau}{x}{\source{\absop{e_L}{e_R}}} \reduces \source{\absop{e_L'}{e_R'}}
-%%}
-%%\end{mathpar}
-%%\subsection{Canonical Forms}
-%%\begin{center}
-%%\begin{tabular}{ | c | c | }
-%%  \hline            
-%%  \textbf{Kind} & \textbf{Canonical Form(s)} \\
-%%  \hline
-%%  $\kappa_1 \rightarrow \kappa_2$ & $\kfn{x}{\kappa}{\tau}$ \\
-%%  $\star$ & $\typedefstd$, $\err$ \\
-%%  $\sfype{\intt}$ & $\evarCtx[\intlit]$ \\
-%%  $\sfype{\tvarCtx_{L} \times \tvarCtx_{R}}$ & $\evarCtx[(\evarCtx_{L}, \evarCtx_{R})]$ where $\evarCtx[\evarCtx_{L}]$ is a canonical form for $\sfype{\tvarCtx_{L}}$ \\ & and $\evarCtx[\evarCtx_{R}]$ is a canonical form for $\sfype{\tvarCtx_{R}}$.\\
-%%  $\sfype{\otimes}$ & $\evarCtx[\fault]$\\
-%%  $\parsetree$ & $\numtreestd$, $\newtreestd$, $\optreestd$\\
-%%  $\sourcet$ & $\source{e}$\\
-%%  \hline  
-%%\end{tabular}
-%%\end{center}
+%%%\begin{mathpar}
+%%%\intlitT
+%%%~~~~
+%%%\addT
+%%%\end{mathpar}\begin{mathpar}
+%%%\tupleT\end{mathpar}\begin{mathpar}
+%%%\fstT
+%%%~~~~
+%%%\sndT
+%%%~~~~
+%%%\faultT
+%%%\end{mathpar}
+%%%\subsection{Metalanguage Statics}
+%%%Type-level terms, $\tau$, form a simply typed lambda calculus. The types of the metalanguage, $\kappa$, are called {\it kinds}. 
+%%%Statics take the form $\entails{\ctx}{\tau \sim \kappa}$, where the context, $\ctx$, operates as a set of assignments of kinds to variables and $\ctx \uplus x \sim \tau$ replaces any previous assignment for $x$ in $\ctx$ with $x\sim\tau$.
+%%%\paragraph{Standard Terms}
+%%%\newcommand{\axiom}[1]{\inferrule{ }{#1}}
+%%%\begin{mathpar}
+%%%\varT~~~~
+%%%\kfnIT\\
+%%%\kfnET
+%%%\end{mathpar}
+%%%For convenience, a simple ${\sf let}$ form is included.
+%%%\begin{mathpar}
+%%%\letT
+%%%\end{mathpar}
+%%%The ${\sf if}$ form checks whether two terms are $\beta$-equivalent. The two branches must evaluate to terms of the same kind.
+%%%\begin{mathpar}
+%%%\ifT
+%%%\end{mathpar}
+%%%The remaining statics are described in Section 3. We list them again, and give each a name, below.
+%%%\paragraph{Type Construction}
+%%%\begin{mathpar}
+%%%\typeT\\
+%%%\errT
+%%%\end{mathpar}
+%%%\paragraph{Code Generation Primitives}
+%%%\begin{mathpar}
+%%%\ilrepT\\
+%%%\iladdT\\
+%%%\iltupleT\\
+%%%\ilfstT~~~~
+%%%\ilsndT
+%%%\end{mathpar}
+%%%
+%%%\paragraph{Source Parsing Rules}
+%%%\begin{mathpar}
+%%%\numtreePT~~~~\newtreePT\\
+%%%\optreePT
+%%%\end{mathpar}
+%%%\paragraph{Syntax Tree Representations}
+%%%\begin{mathpar}
+%%%\numtreeT\\\newtreeT\end{mathpar}{\begin{mathpar}\optreeT\\\visitT
+%%%\end{mathpar}
+%%%\paragraph{Source Program Representation}
+%%%\begin{mathpar}
+%%%\sourcetT
+%%%\end{mathpar}
+%%%\subsection{Metalanguage Dynamics}
+%%%We now list the dynamics of the metalanguage.
+%%%\paragraph{Standard Terms}
+%%%\begin{mathpar}
+%%%\inferrule{ }{
+%%%	\final{\kfn{x}{\kappa}{\tau}}
+%%%}
+%%%~~~~
+%%%\inferrule{
+%%%	\tau_0 \reduces \tau_0'
+%%%}{
+%%%	\tau_0~\tau_1 \reduces \tau_0'~\tau_1
+%%%}
+%%%\\
+%%%\inferrule{ }{
+%%%	(\kfn{x}{\kappa}{\tau})~\tau_0 \reduces \subst{\tau_0}{x}{\tau}
+%%%}
+%%%~~~~
+%%%\inferrule{ }{
+%%%	\klet{x}{\tau_0}{\tau_1} \reduces \subst{\tau_0}{x}{\tau_1}
+%%%}\\
+%%%\inferrule{\tau_0 \reduces \tau_0'}{
+%%%	\ifeq{\tau_0=\tau_1}{\tau_\true}{\tau_\false} \reduces \ifeq{\tau_0' = \tau_1}{\tau_\true}{\tau_\false}
+%%%}\\
+%%%\inferrule{\final{\tau_0}\\\tau_1 \reduces \tau_1'}{
+%%%	\ifeq{\tau_0=\tau_1}{\tau_\true}{\tau_\false} \reduces \ifeq{\tau_0 = \tau_1'}{\tau_\true}{\tau_\false}
+%%%}\\
+%%%\inferrule{\final{\tau_0}\\\final{\tau_1}\\\tau_0 = \tau_1}{
+%%%	\ifeq{\tau_0=\tau_1}{\tau_\true}{\tau_\false} \reduces \tau_\true
+%%%}\\
+%%%\inferrule{\final{\tau_0}\\\final{\tau_1}\\\tau_0 \neq \tau_1}{
+%%%	\ifeq{\tau_0=\tau_1}{\tau_\true}{\tau_\false} \reduces \tau_\false
+%%%}
+%%%\end{mathpar}
+%%%$\tau_0 = \tau_1$ and $\tau_0 \neq \tau_1$ denote syntactic equality and inequality.
+%%%\paragraph{Type Construction}
+%%%\begin{mathpar}
+%%%\inferrule{ }{
+%%%	\final{\typedefstd}
+%%%}
+%%%~~~~
+%%%\inferrule{ }{
+%%%	\final{\err}
+%%%}
+%%%\end{mathpar}
+%%%\paragraph{Code Generation Primitives}
+%%%\begin{mathpar}
+%%%\inferrule{ }{
+%%%	\final{\evarCtx[\evarCtx]}
+%%%}
+%%%~~~~
+%%%\inferrule{
+%%%	\tau_L \reduces \tau_L'
+%%%}{
+%%%	\evarCtx_{+}(\tau_L, \tau_R) \reduces \evarCtx_{+}(\tau_L', \tau_R)
+%%%}\end{mathpar}\begin{mathpar}
+%%%\inferrule{
+%%%	\tau_R \reduces \tau_R'	
+%%%}{
+%%%	\evarCtx_{+}(\evarCtx[\evarCtx_L], \tau_R) \reduces \evarCtx_{+}(\evarCtx[\evarCtx_L], \tau_R')
+%%%}\end{mathpar}\begin{mathpar}
+%%%\inferrule{ }{
+%%%	\evarCtx_{+}(\evarCtx[\evarCtx_L], \evarCtx[\evarCtx_R]) \reduces \evarCtx[\evarCtx_L + \evarCtx_R]
+%%%}\end{mathpar}\begin{mathpar}
+%%%\inferrule{
+%%%	\tau_L \reduces \tau_L'
+%%%}{
+%%%	\evarCtx_{\times}(\tau_L, \tau_R) \reduces \evarCtx_{\times}(\tau_L', \tau_R)
+%%%}\end{mathpar}\begin{mathpar}
+%%%\inferrule{
+%%%	\tau_R \reduces \tau_R'	
+%%%}{
+%%%	\evarCtx_{\times}(\evarCtx[\evarCtx_L], \tau_R) \reduces \evarCtx_{\times}(\evarCtx[\evarCtx_L], \tau_R')
+%%%}\end{mathpar}\begin{mathpar}
+%%%\inferrule{ }{
+%%%	\evarCtx_{\times}(\evarCtx[\evarCtx_L], \evarCtx[\evarCtx_R]) \reduces \evarCtx[(\evarCtx_L, \evarCtx_R)]
+%%%}\end{mathpar}\begin{mathpar}
+%%%\inferrule{\tau_\times \reduces \tau_\times'}{
+%%%	\evarCtxfst{\tau_\times} \reduces \evarCtxfst{\tau_\times'}
+%%%}
+%%%~~~~
+%%%\inferrule{ }{
+%%%	\evarCtxfst{\evarCtx[(\evarCtx_L, \evarCtx_R)]} \reduces \evarCtx[\evarCtx_L]
+%%%}\end{mathpar}\begin{mathpar}
+%%%\inferrule{\tau_\times \reduces \tau_\times'}{
+%%%	\evarCtxsnd{\tau_\times} \reduces \evarCtxsnd{\tau_\times'}
+%%%}
+%%%~~~~
+%%%\inferrule{ }{
+%%%	\evarCtxsnd{\evarCtx[(\evarCtx_L, \evarCtx_R)]} \reduces \evarCtx[\evarCtx_R]
+%%%}
+%%%\end{mathpar}
+%%%\paragraph{Syntax Tree Representations}
+%%%\begin{mathpar}
+%%%\inferrule{
+%%%	\tau_n \reduces \tau_n'
+%%%}{
+%%%	\numtree{\tau_n} \reduces \numtree{\tau_n'}
+%%%}~~~~
+%%%\axiom{\final{\numtree{\evarCtx[\intlit]}}}\end{mathpar}\begin{mathpar}
+%%%\inferrule{
+%%%	\tau_\star \reduces \tau_\star'
+%%%}{
+%%%	\newtreestd \reduces \newtree{\tau_\star'}{\tau_\pi}
+%%%}~~~~
+%%%\inferrule{
+%%%	\final{\tau_\star}\\
+%%%	\tau_\pi \reduces \tau_\pi'
+%%%}{
+%%%	\newtreestd \reduces \newtree{\tau_\star}{\tau_\pi'}
+%%%}\end{mathpar}\begin{mathpar}
+%%%\inferrule{
+%%%	\final{\tau_\star}\\
+%%%	\final{\tau_\pi}
+%%%}{
+%%%	\final{\newtreestd}
+%%%}\end{mathpar}\begin{mathpar}
+%%%\inferrule{
+%%%	\tauL \reduces \tauL'
+%%%}{
+%%%	\optreestd \reduces \optree{\tauL'}{\tauR}
+%%%}~~~~
+%%%\inferrule{
+%%%	\final{\tauL}\\
+%%%	\tauR \reduces \tauR'
+%%%}{
+%%%	\optreestd \reduces \optree{\tauL}{\tauR'}
+%%%}\end{mathpar}\begin{mathpar}
+%%%\inferrule{
+%%%	\final{\tauL}\\
+%%%	\final{\tauR}
+%%%}{
+%%%	\final{\optreestd}
+%%%}
+%%%\end{mathpar}
+%%%\begin{mathpar}
+%%%\inferrule{
+%%%	\taupi \reduces \taupi'
+%%%}{
+%%%	\kvisitstd \reduces \kvisit{\taupi'}{\tau_{\#}}{\tauadd}{\taunew}
+%%%}\end{mathpar}\begin{mathpar}
+%%%\axiom{\kvisit{\numtreestd}{\tau_{\#}}{\tauadd}{\taunew} \reduces \tau_{\#}~\tau_n}\end{mathpar}\begin{mathpar}
+%%%\axiom{\kvisit{\newtreestd}{\tau_{\#}}{\tauadd}{\taunew} \reduces \taunew~\tau_\star~\taupi}\end{mathpar}\begin{mathpar}
+%%%\axiom{\kvisit{\optreestd}{\tau_{\#}}{\tauadd}{\taunew} \reduces \tauadd~\tauL~\tauR}\end{mathpar}
+%%%\paragraph{Source Program Representations}
+%%%\begin{mathpar}
+%%%\axiom{\final{\source{\intlit}}}\\
+%%%\inferrule{
+%%%	\tau \reduces \tau'
+%%%}{
+%%%	\source{\newobj{\tau}{e}} \reduces \source{\newobj{\tau'}{e}}
+%%%}\\
+%%%\inferrule{
+%%%	\final{\tau}\\
+%%%	\source{e} \reduces \source{e'}
+%%%}{
+%%%	\source{\newobj{\tau}{e}} \reduces \source{\newobj{\tau}{e'}}
+%%%}\\
+%%%\inferrule{
+%%%	\final{\tau}\\
+%%%	\final{\source{e}}
+%%%}{
+%%%	\final{\source{\newobj{\tau}{e}}}
+%%%}\\
+%%%\inferrule{
+%%%	\source{e_L} \reduces \source{e_L'}
+%%%}{
+%%%	\source{\absop{e_L}{e_R}} \reduces \source{\absop{e_L'}{e_R}}
+%%%}\\
+%%%\inferrule{
+%%%	\final{\source{e_L}}\\
+%%%	\source{e_R} \reduces \source{e_R'}
+%%%}{
+%%%	\source{\absop{e_L}{e_R}} \reduces \source{\absop{e_L}{e_R'}}
+%%%}\\
+%%%\inferrule{
+%%%	\final{\source{e_L}}\\
+%%%	\final{\source{e_R}}
+%%%}{
+%%%	\final{\source{\absop{e_L}{e_R}}}
+%%%}
+%%%\end{mathpar}
+%%%\paragraph{Substitution Dynamics}
+%%%\subparagraph{Standard Terms}
+%%%\begin{mathpar}
+%%%\inferrule{ }{
+%%%	\subst{\tau}{x}{x} \reduces \tau
+%%%}
+%%%~~~~
+%%%\inferrule{x \neq y}{
+%%%	\subst{\tau}{x}{y} \reduces y
+%%%}\end{mathpar}\begin{mathpar}
+%%%\inferrule{ }{
+%%%	\subst{\tau}{x}{\kfn{x}{\kappa}{\tau_0}} \reduces \kfn{x}{\kappa}{\tau_0}
+%%%}\end{mathpar}\begin{mathpar}
+%%%\inferrule{
+%%%	x \neq y\\
+%%%	\subst{\tau}{x}{\tau_0} \reduces \tau_0'	
+%%%}{
+%%%	\subst{\tau}{x}{\kfn{y}{\kappa}{\tau_0}} \reduces \kfn{y}{\kappa}{\tau_0'}
+%%%}\end{mathpar}\begin{mathpar}
+%%%\inferrule{
+%%%	\subst{\tau}{x}{\tau_0} \reduces \tau_0'\\
+%%%	\subst{\tau}{x}{\tau_1} \reduces \tau_1'
+%%%}{
+%%%	\subst{\tau}{x}{\tau_0~\tau_1} \reduces \tau_0'~\tau_1'
+%%%}\end{mathpar}\begin{mathpar}
+%%%\inferrule{
+%%%	\subst{\tau}{x}{\tau_0} \reduces \tau_0'\\
+%%%}{
+%%%	\subst{\tau}{x}{\klet{x}{\tau_0}{\tau_1}} \reduces \klet{x}{\tau_0'}{\tau_1}
+%%%}\end{mathpar}\begin{mathpar}
+%%%\inferrule{
+%%%	x \neq y\\
+%%%	\subst{\tau}{x}{\tau_0} \reduces \tau_0'\\
+%%%	\subst{\tau}{x}{\tau_1} \reduces \tau_1'
+%%%}{
+%%%	\subst{\tau}{x}{\klet{y}{\tau_0}{\tau_1}} \reduces \klet{y}{\tau_0'}{\tau_1'}
+%%%}\end{mathpar}\begin{mathpar}
+%%%\inferrule{
+%%%	\subst{\tau}{x}{\tau_0} \reduces \tau_0'~~~~
+%%%	\subst{\tau}{x}{\tau_1} \reduces \tau_1'~~~~
+%%%	\subst{\tau}{x}{\tau_\true} \reduces \tau_\true'~~~~
+%%%	\subst{\tau}{x}{\tau_\false} \reduces \tau_\false'
+%%%}{
+%%%	\subst{\tau}{x}{\ifeq{\tau_0 = \tau_1}{\tau_\true}{\tau_\false}} \reduces \ifeq{\tau_0' = \tau_1'}{\tau_\true'}{\tau_\false'}
+%%%}\end{mathpar}
+%%%\subparagraph{Type Construction}
+%%%\begin{mathpar}
+%%%\inferrule{
+%%%	\subst{\tau}{x}{\taunewg} \reduces \hat{\tau}_{\sf new}^\evarCtx\\
+%%%	\subst{\tau}{x}{\tauaddg} \reduces \hat{\tau}_\diamond^\star\\
+%%%	\subst{\tau}{x}{\tauaddt} \reduces \hat{\tau}_\diamond^\evarCtx
+%%%}{
+%%%	\subst{\tau}{x}{\typedefstd} \reduces \subst{\tau}{x}{\typedef{a}{\hat{\tau}_{\sf new}^\evarCtx}{\hat{\tau}_\diamond^\star}{\hat{\tau}_\diamond^\evarCtx}{\tvarCtxmap}{f}{g}}
+%%%}\end{mathpar}\begin{mathpar}
+%%%\inferrule{ }{
+%%%	\subst{\tau}{x}{\err} \reduces \err
+%%%}\end{mathpar}
+%%%\subparagraph{Code Generation Primitives}
+%%%\begin{mathpar}
+%%%\inferrule{ }{
+%%%	\subst{\tau}{x}{\evarCtx[\evarCtx]} \reduces \evarCtx[\evarCtx]
+%%%}\end{mathpar}\begin{mathpar}
+%%%\inferrule{
+%%%	\subst{\tau}{x}{\tau_L} \reduces \tau_L'\\
+%%%	\subst{\tau}{x}{\tau_R} \reduces \tau_R'
+%%%}{
+%%%	\subst{\tau}{x}{\evarCtx_{+}(\tau_L, \tau_R)} \reduces \evarCtx_{+}(\tau_L', \tau_R')
+%%%}\end{mathpar}\begin{mathpar}
+%%%\inferrule{
+%%%	\subst{\tau}{x}{\tau_L} \reduces \tau_L'\\
+%%%	\subst{\tau}{x}{\tau_R} \reduces \tau_R'
+%%%}{
+%%%	\subst{\tau}{x}{\evarCtx_{\times}(\tau_L, \tau_R)} \reduces \evarCtx_{\times}(\tau_L', \tau_R')
+%%%}\end{mathpar}\begin{mathpar}
+%%%\inferrule{
+%%%	\subst{\tau}{x}{\tau_\times} \reduces \tau_\times'
+%%%}{
+%%%	\subst{\tau}{x}{\evarCtxfst{\tau_\times}} \reduces \evarCtxfst{\tau_\times'}
+%%%}
+%%%~~~~
+%%%\inferrule{
+%%%	\subst{\tau}{x}{\tau_\times} \reduces \tau_\times'
+%%%}{
+%%%	\subst{\tau}{x}{\evarCtxsnd{\tau_\times}} \reduces \evarCtxsnd{\tau_\times'}
+%%%}\end{mathpar}
+%%%\subparagraph{Syntax Tree Representations}
+%%%\begin{mathpar}
+%%%\inferrule{
+%%%	\subst{\tau}{x}{\tau_n} \reduces \tau_n'
+%%%}{
+%%%	\subst{\tau}{x}{\numtreestd} \reduces \numtree{\tau_n'}
+%%%}
+%%%~~~~
+%%%\inferrule{
+%%%	\subst{\tau}{x}{\tau_\star} \reduces \tau_\star'\\
+%%%	\subst{\tau}{x}{\tau_\pi} \reduces \tau_\pi'
+%%%}{
+%%%	\subst{\tau}{x}{\newtreestd} \reduces \newtree{\tau_\star'}{\tau_\pi'}
+%%%}\end{mathpar}\begin{mathpar}
+%%%\inferrule{
+%%%	\subst{\tau}{x}{\tau_L} \reduces \tau_L'\\
+%%%	\subst{\tau}{x}{\tau_R} \reduces \tau_R'
+%%%}{
+%%%	\subst{\tau}{x}{\optreestd} \reduces \optree{\tau_L'}{\tau_R'}
+%%%}\end{mathpar}\begin{mathpar}
+%%%\inferrule{
+%%%	\subst{\tau}{x}{\tau_\pi} \reduces \tau_\pi'~~~~
+%%%	\subst{\tau}{x}{\tau_{\#}} \reduces \tau_{\#}'\\
+%%%	\subst{\tau}{x}{\tau_{\sf new}} \reduces \tau_{\sf new}'~~~~
+%%%	\subst{\tau}{x}{\tau_\diamond} \reduces \tau_\diamond
+%%%}{
+%%%	\subst{\tau}{x}{\kvisit{\tau_\pi}{\tau_{\#}}{\tau_\diamond}{\tau_{\sf new}}} \reduces
+%%%	\kvisit{\tau_\pi'}{\tau_{\#}'}{\tau_\diamond'}{\tau_{\sf new}'}
+%%%}\end{mathpar}
+%%%\subparagraph{Source Program Representations}
+%%%\begin{mathpar}
+%%%\inferrule{ }{
+%%%	\subst{\tau}{x}{\source{\intlit}} \reduces \source{\intlit}
+%%%}\end{mathpar}\begin{mathpar}
+%%%\inferrule{
+%%%	\subst{\tau}{x}{\tau_\star} \reduces \tau_\star'\\
+%%%	\subst{\tau}{x}{\source{e}} \reduces \source{e'}
+%%%}{
+%%%	\subst{\tau}{x}{\source{\newobj{\tau_\star}{e}}} \reduces \source{\newobj{\tau_\star'}{e'}}
+%%%}\end{mathpar}\begin{mathpar}
+%%%\inferrule{
+%%%	\subst{\tau}{x}{\source{e_L}} \reduces \source{e_L'}\\
+%%%	\subst{\tau}{x}{\source{e_R}} \reduces \source{e_R'}
+%%%}{
+%%%	\subst{\tau}{x}{\source{\absop{e_L}{e_R}}} \reduces \source{\absop{e_L'}{e_R'}}
+%%%}
+%%%\end{mathpar}
+%%%\subsection{Canonical Forms}
+%%%\begin{center}
+%%%\begin{tabular}{ | c | c | }
+%%%  \hline            
+%%%  \textbf{Kind} & \textbf{Canonical Form(s)} \\
+%%%  \hline
+%%%  $\kappa_1 \rightarrow \kappa_2$ & $\kfn{x}{\kappa}{\tau}$ \\
+%%%  $\star$ & $\typedefstd$, $\err$ \\
+%%%  $\sfype{\intt}$ & $\evarCtx[\intlit]$ \\
+%%%  $\sfype{\tvarCtx_{L} \times \tvarCtx_{R}}$ & $\evarCtx[(\evarCtx_{L}, \evarCtx_{R})]$ where $\evarCtx[\evarCtx_{L}]$ is a canonical form for $\sfype{\tvarCtx_{L}}$ \\ & and $\evarCtx[\evarCtx_{R}]$ is a canonical form for $\sfype{\tvarCtx_{R}}$.\\
+%%%  $\sfype{\otimes}$ & $\evarCtx[\fault]$\\
+%%%  $\parsetree$ & $\numtreestd$, $\newtreestd$, $\optreestd$\\
+%%%  $\sourcet$ & $\source{e}$\\
+%%%  \hline  
+%%%\end{tabular}
+%%%\end{center}
 \end{document}

### 2014-03-15 thesis-proposal/background.tex
removed stuff

@@
 \cite{activelibraries, active-libraries-thesis} to describe ``libraries that take an active role in compilation, rather than being passive collections of subroutines''.
-The authors suggested a number of reasons libraries might benefit by being able to influence the programming system at compile-time or edit-time, including high-level program optimization, checking programs for correctness against specialized criteria, reporting domain-specific errors and warnings, and ``rendering domain-specific textual and non-textual program representations and for interacting with such representations'' (anticipating interactions between libraries and tools other than just the compiler).
+The authors suggested a number of reasons libraries might benefit from being able to influence the programming system directly, including high-level program optimization, checking programs for correctness against specialized criteria, reporting domain-specific errors and warnings, and ``rendering domain-specific textual and non-textual program representations and for interacting with such representations'' (anticipating interactions between libraries and tools other than just the compiler).
 
@@
 More powerful and direct compile-time \emph{term rewriting mechanisms} available in some languages can also be used for optimization, as well as for introducing specialized error checking logic and implementing new abstractions.
-These mechanisms  suffer from problems of composability, safety and expressiveness.
-For example, compile-time macros, such as those in MetaML \cite{Sheard:1999:UMS}, Template Haskell \cite{SheardPeytonJones:Haskell-02} and Scala \cite{scala-macros}, take full control over all of the code that they enclose.
-This can be problematic, however, as outer macros can interfere with the functionality of inner macros.
+For example, typed macros, such as those in MetaML \cite{Sheard:1999:UMS}, Template Haskell \cite{SheardPeytonJones:Haskell-02} and Scala \cite{scala-macros}, take full control over all of the code that they enclose.
+This can be problematic, however, as outer macros can interfere with inner macros.
 Moreover, once a value escapes a macro's scope, there is no way to rely on the guarantees and features that were available within its scope, because the output of a macro is simply a term in the underlying language (a problem fundamentally related to the problem of relying on a common intermediate language,  described in Section \ref{external-approaches}).
-Thus, macros can be used to automate code generation, but not to globally extend the syntax or static guarantees of a language.
+Thus, macros represent at best a partial solution: they can be used to automate code generation, but not to globally extend the syntax or static guarantees of a language.
 It can also be difficult to reason about the semantics of code when any number of enclosing macros may be manipulating it, and to build tools that operate robustly in their presence.
 
-Some term rewriting systems replace the explicitly delimited scoping of macros with global pattern-based dispatch.
-Xroma (pronounced ``Chroma''), for example, is designed around active libraries and allows users to insert custom rewriting passes into the compiler from within libraries \cite{activelibraries}.
-Similarly, the derivative of Haskell implemented by the Glasgow Haskell Compiler (GHC) allows providers to introduce custom compile-time term rewriting logic if an appropriate flag is passed in \cite{jones2001playing}.
-In both cases, the user-defined logic can dispatch on arbitrary patterns of code throughout the component or program the extension is activated within, so these mechanisms permit non-local extensions to the static and dynamic semantics.
+Some term rewriting systems replace the delimited scoping of macros with global pattern-based dispatch.
+Xroma (pronounced ``Chroma''), for example, allows users to insert custom rewriting passes into the compiler from within libraries \cite{activelibraries}.
+Similarly, the Glasgow Haskell Compiler (GHC) allows providers to introduce custom compile-time term rewriting logic if an appropriate flag is passed in \cite{jones2001playing}.
+In both cases, the user-defined logic can dispatch on arbitrary patterns of code throughout the component or program the extension is activated within.
 But libraries containing such global rewriting logic cannot be safely composed because two different libraries may attempt to rewrite the same piece of code differently.

### 2014-03-15 thesis-proposal/parsing.tex
removed stuff

@@
 % !TEX root = omar-thesis-proposal.tex
-\lstset{tabsize=2, 
-basicstyle=\ttfamily\fontsize{8pt}{1em}\selectfont, 
-commentstyle=\itshape\ttfamily, 
-stringstyle=\ttfamily,
-numbers=left, numberstyle=\scriptsize\color{gray}\ttfamily, language=ML,moredelim=[il][\sffamily]{?},mathescape=true,showspaces=false,showstringspaces=false,xleftmargin=15pt, morekeywords=[1]{tyfam,opfam,let,fn,val,def,casetype,objtype,metadata,of,*,datatype,new,valAST},deletekeywords={for,double,in},classoffset=0,belowskip=\smallskipamount,
-moredelim=**[is][\color{cyan}]{SSTR}{ESTR},
-moredelim=**[is][\color{OliveGreen}]{SHTML}{EHTML},
-moredelim=**[is][\color{purple}]{SCSS}{ECSS},
-moredelim=**[is][\color{brown}]{SSQL}{ESQL},
-moredelim=**[is][\color{orange}]{SCOLOR}{ECOLOR},
-moredelim=**[is][\color{magenta}]{SPCT}{EPCT}, 
-moredelim=**[is][\color{gray}]{SNAT}{ENDNAT}, 
-moredelim=**[is][\color{blue}]{SURL}{EURL},
-moredelim=**[is][\color{SeaGreen}]{SQT}{EQT},
-moredelim=**[is][\color{Periwinkle}]{SGRM}{EGRM},
-moredelim=**[is][\color{YellowGreen}]{SID}{EID},
-moredelim=**[is][\color{Sepia}]{SUS}{EUS}
-}
-\lstloadlanguages{Java,VBScript,XML,HTML}
-\let\li\lstinline
-
 \section{Type-Specific Languages}\label{aparsing}
-General-purpose abstraction mechanisms like those built into modern languages are semantically quite flexible.
-By using a general-purpose abstraction mechanism to define an abstraction, one immediately benefits from a body of useful operations, established reasoning principles, well-optimized implementations and tool support.
-For example, lists can be encoded using a more general mechanism for defining  inductive datatypes.
+General-purpose constructs like those built into modern languages are, semantically, quite powerful.
+By using a general-purpose construct to define a data structure, one immediately benefits from a body of useful operators, established reasoning principles, well-optimized implementations and tool support.
+For example, lists can be encoded using a more general mechanism for defining polymorphic recursive sum types.
 Intuitively, a list can either be empty, or be broken down into a \emph{head} element and a \emph{tail}, another list.
-In an ML-like language, the polymorphic list type is declared:
+In an ML-like language, the polymorphic list type would be defined:
 \begin{lstlisting}[numbers=none]
@@
 \end{lstlisting}
-By encoding lists in this way, we can immediately reason about them by structural induction and examine them by pattern matching.
+By defining lists in this way, we can immediately reason about list values by structural induction and examine them by pattern matching.
 %The programmer only needs to provide an encoding of the structure of lists; the semantics and implementation are filled in by the general-purpose abstraction mechanism.
-
-While these operations and reasoning principles can be quite useful, the associated general-purpose syntax can sometimes be a liability.
-For example, few would claim that writing a list of numbers as a sequence of \li{Cons} cells is convenient:
+While this is quite useful, the associated general-purpose notation is not. For example, few would claim that writing a list of numbers as a sequence of \li{Cons} cells is convenient:
 \begin{lstlisting}[numbers=none]
@@
 
-Because lists are a common data structure, many languages include \emph{literal syntax} for introducing them, e.g.
+Because lists are a common data structure, many languages include \emph{literal notations} for introducing them, e.g.
 \li{[1, 2, 3, 4]}.
 This notation is semantically equivalent to the general-purpose notation shown above, but brings cognitive benefits by drawing attention to the content of the list, rather than the nature of the encoding.
-To use terminology from the literature on the cognitive dimensions of notations \cite{green1996usability}, it is more \emph{terse}, \emph{visible} and \emph{maps more closely} to the intuitive notion of a list.
+Using terminology from Green's cognitive dimensions of notations \cite{green1996usability}, it is more \emph{terse}, \emph{visible} and \emph{maps more closely} to the intuitive notion of a list.
+Stoy, in writing about the value of good notation, says \cite{stoybook}:
+\begin{quote}A good notation thus conceals much of the inner workings behind suitable abbreviations, while allowing us to consider it in more detail if we require: matrix and tensor notations provide further good examples of this.
+It may be summed up in the saying: 'A notation is important for what it leaves out.'\end{quote}
 
-Although list, number and string literals are nearly ubiquitous features of modern languages, some languages also  provide specialized notation for other common data structures (like maps and sets), data formats (like XML and JSON), query languages (like regular expressions and SQL), markup languages (like HTML) and many other types of data.
-For example, a language with built-in notation for HTML and SQL, with type-safe \emph{interpolation} of host language terms within curly braces, might define:
+Although list, number and string literals are nearly ubiquitous features of modern languages, some languages also  provide specialized notation for other common data structures (like maps, sets, vectors and matrices), data formats (like XML and JSON), query languages (like regular expressions and SQL), markup languages (like HTML and Markdown) and many other types of data.
+For example, a language with built-in notation for HTML and SQL, supporting type-safe \emph{interpolation}\todo{splicing?} via curly braces, might define:
 \begin{lstlisting}
@@
 When a specialized notation like this is not available, but the equivalent general-purpose notation is too cognitively demanding for comfort, developers typically turn to run-time mechanisms to make constructing data structures more convenient.
-Among the most common strategies in these situations is to simply use a string representation that is parsed at run-time.
-Developers across language paradigms frequently write examples like the above as:
+Among the most common strategies across language paradigms in these situations is to simply use a string representation that is parsed at run-time.% Developers across language paradigms frequently write examples like the above as:
 \begin{lstlisting}
@@
 Although developers are cautioned to sanitize their input, it can be difficult to verify that this was done correctly throughout a codebase.
-The most straightforward way to avoid these problems today is to insist on structured representations, despite their inconvenience.
+The best way to avoid these problems today is to avoid strings and insist on structured representations, despite the inconvenient notation.
 
-Unfortunately, it does not appear that this is working.
-Situations like this, where maintaining strong correctness, performance and security guarantees entails significant syntactic overhead, causing developers to turn to worse solutions that are more convenient, are quite common.
+Unfortunately, situations like this, where maintaining strong correctness, performance and security guarantees entails significant syntactic overhead, causing developers to turn to worse solutions that are more convenient, are quite common.
 To emphasize this, let us return to our running example of pattern literals.
@@
 \end{lstlisting}
-This is clearly more cognitively demanding, both when authoring the regular expression and when reading it.
-Among the most common strategies in these situations, for users of both object-oriented and functional languages, is to simply use a string representation that is parsed at run-time.
+This is clearly more cognitively demanding, both when writing the regular expression and when reading it.
+Among the most common strategies in these situations, for users of any kind of language, is again to simply use a string representation that is parsed at run-time.
 \begin{lstlisting}
@@
 So although languages that allow users to introduce new syntax from within libraries appear to hold promise for the reasons described above, providing this form of extensibility is non-trivial because there is no longer a central designer making decisions about such ambiguities.
-In most existing related work, the burden of resolving ambiguities falls to the clients using conflicting extensions.
+In most existing related work, the burden of resolving ambiguities falls to clients who have the misfortune of importing conflicting extensions.
 For example, SugarJ \cite{erdweg2011sugarj} and other extensible languages generated by Sugar* \cite{erdweg2013framework} allow providers to extend the base syntax of the host language with new forms, like set and map literals.
@@
 This is both difficult to do, requiring a reasonably thorough understanding of the underlying parser technology (in Sugar*, generalized LR parsing) and increases the cognitive load of using the conflicting notations (e.g.
-both sets and dictionaries) in the same file.
+both sets and dictionaries) in the same file because disambiguation tokens must be used.
 These kinds of conflicts occur in a variety of circumstances: HTML and XML, different variants of SQL, JSON literals and dictionaries, or simply different implementations (``desugarings'') of the same specialized syntax (e.g.
@@
 This strategy avoids the problem of conflicting syntax, because neither the base language nor TSLs are ever extended directly.
-It also permits semantic flexibility -- the meaning of a form like \verb|{ }| can differ depending  on its type, so it is safe to use it for empty sets, dictionaries and other data structures, like JSON literals.
+It also permits semantic flexibility -- the meaning of a form like \verb|{ }| can differ depending  on its type, so it is safe to use it for empty sets, maps and other data structures, like JSON literals.
 This frees these common notations from being tied to the variant of a  data structure built into a language's standard library, which sometimes does not provide the exact semantics that a programmer needs (for example, Python dictionaries do not preserve key insertion order).
-%\begin{lstlisting}
-%let empty_set : Set = { }
-%let empty_dict : Dict = { }
-%let empty_json : JSON = { }
-%\end{lstlisting}
-\subsection{Wyvern}
-We develop our work as a variant of a new programming language being developed by our group called Wyvern \cite{Nistor:2013:WST:2489828.2489830}.
-To allow us to focus on the essence of our proposal, the variant of Wyvern we will describe in this thesis is simpler than the variant previously described: it is purely functional (there are no effects other than non-termination) and it does not enforce a uniform access principle for objects (fields can be accessed directly).
-Objects are thus a simple variant of records equipped with simple methods (functions that are automatically given a self-reference) for convenience.
-We also add recursive labeled sum types, which we call \emph{case types}, that are quite similar to datatypes in ML.
-One can refer to the version of the language described in this thesis as \emph{TSL Wyvern}.
-
-\subsection{Example: Web Search}
-We begin in Fig.
-\ref{f-example} with an example showing several different TSLs being used to define a fragment of a web application showing search results from a database.
-Note that for clarity of presentation, we color each character  according to the TSL it is governed by.
 \begin{figure}[t]
@@
 SCSS            body { background-image: url(%ECSSbgImageSCSS%) }
-            #search { background-color: %ECSS`SCOLOR#aabbccECOLOR`.darken(SPCT20pctEPCT)SCSS% }
+            #search { background-color: %ECSS`SCOLOR#aabbccECOLOR`SCSS% }
 ECSSSHTML        :body
-          :h1 Results for {EHTMLsearchQuerySHTML}
+          :h1 Results for :{EHTMLsearchQuerySHTML}
           :div[id="search"]
-            Search again: {EHTMLSearchBox($\texttt{"}$SSTRGo!ESTR$\texttt{"}$)SHTML}
-          {EHTML (* fmt_results : DB * SQLQuery * Nat * Nat -> HTML *)
+            Search again: :{EHTMLSearchBox($\texttt{"}$SSTRGo!ESTR$\texttt{"}$)SHTML}
+          :wyvEHTML (* fmt_results : DB * SQLQuery * Nat * Nat -> HTML *)
             fmt_results(db, ~, SNAT10ENDNAT, page)
               SSQLSELECT * FROM products WHERE {ESQLsearchQuerySSQL} in titleESQL
-          SHTML}
 \end{lstlisting}
@@
 \end{figure}
+%\begin{lstlisting}
+%let empty_set : Set = { }
+%let empty_dict : Dict = { }
+%let empty_json : JSON = { }
+%\end{lstlisting}
+\subsection{Wyvern}
+We develop our work as a variant of a new programming language being developed by our group called Wyvern \cite{Nistor:2013:WST:2489828.2489830}.
+To allow us to focus on the essence of our proposal, the variant of Wyvern we will describe in this thesis is simpler than the variant previously described: it is purely functional (there are no effects other than non-termination) and it does not enforce a uniform access principle for objects (fields can be accessed directly).
+Objects can thus be thought of as functional records equipped with simple methods (functions that are automatically given a self-reference) for convenience.
+We also add recursive labeled sum types, which we call \emph{case types}, that are quite similar to datatypes in ML.
+One can refer to the version of the language described in this thesis as \emph{TSL Wyvern}.
+
+\subsection{Example: Web Search}
+We begin in Fig.
+\ref{f-example} with an example showing several different TSLs being used to define a fragment of a web application showing search results from a database.
+Note that for clarity of presentation, we color each character  according to the TSL it is governed by.% Black represents the base language.
+
 \subsection{Inline Literals}
 \begin{figure}[t]
-\begin{lstlisting}
+\begin{lstlisting}[numbers=none]
 <SURLliteral body here, <inner angle brackets> must be balancedEURL>
@@
 Our first TSL appears on the right-hand side of the variable binding on line 1.
-The variable \li{imageBase} is annotated with its type, \li{URL} (not shown).
-This is an {object type} declaring several fields representing the components of a URL: its protocol, domain name, port, path and so on.
+The variable \li{imageBase} is annotated with its type, \li{URL}.
+This is an {object type} declaring several fields representing the components of a URL: its protocol, domain name, port, path and so on (not shown).
 We could have created a value of type \li{URL} using general-purpose notation:
@@
 let imageBase : URL = new
-  val protocol : URLFrag = $\text{"}$SUShttpEUS$\texttt{"}$
-  val subdomain : URLFrag = $\texttt{"}$SUSimagesEUS$\texttt{"}$
+  val protocol = $\text{"}$SUShttpEUS$\texttt{"}$
+  val subdomain = $\texttt{"}$SUSimagesEUS$\texttt{"}$
   (* ... *)
@@
 This is tedious.
-Instead, because the \li{URL} type has a TSL associated with it, we can instead introduce precisely this value using conventional notation for URLs by placing it in the \emph{body} of a \emph{generic literal}, \li{<SURLimages.example.comEURL>}.
+Because the \li{URL} type has a TSL associated with it, we can instead introduce precisely this value using conventional notation for URLs by placing it in the \emph{body} of a \emph{generic literal}, \li{<SURLimages.example.comEURL>}.
 Any other delimited form in Fig.
@@
 The type annotation on \li{imageBase} implies that this literal's \emph{expected type} is \li{URL}, so the {body} of the literal (the characters between the angle brackets, in blue) will be governed by the \li{URL} TSL during the typechecking phase.
-This TSL will parse the body ({at compile-time}) to produce a Wyvern abstract syntax tree (AST) that explicitly instantiates a new object of type \li{URL} using general-purpose notation.
+This TSL will parse the body ({at compile-time}) to produce a Wyvern abstract syntax tree (AST) that explicitly instantiates a new object of type \li{URL} using general-purpose notation as if the above had been written directly.
+We will return to how this works shortly.
 
-In addition to supporting conventional notation for URLs, this TSL supports \emph{typed interpolation}
-of another Wyvern expression of type \li{URL} to form a larger URL. The interpolated term is delimited by percent signs,
-as seen on line 2 of Fig.
-\ref{f-example}.
-The TSL parses code between percent signs  as a Wyvern expression with expected type \li{URL}, using its AST to construct an AST for the expression as a whole.
-String interpolation
-never occurs. Note that the delimiters that are used to go from Wyvern to a TSL are controlled by Wyvern (those in Fig. \ref{f-delims}) while the TSL controls how to return to Wyvern.
-
+In addition to supporting conventional notation for URLs, this TSL supports \emph{interpolating}
+another Wyvern expression of type \li{URL} to form a larger URL. The interpolated term is delimited by percent signs,
+as seen on line 2 of Fig. \ref{f-example}. The TSL parses code between percent signs  as a Wyvern expression, using its AST to construct an AST for the expression as a whole. String concatenation  
+never occurs. Note that the delimiters used to go from Wyvern to a TSL are controlled by Wyvern  while the TSL controls how to return to Wyvern. 
 %\subsection{General-Purpose Notation for Object and Case Types}
@@
 
+On line 5 of Fig.
+\ref{f-example}, we see a call to a function \li{serve} (not shown) which has type \li{HTML -> Unit}.
+Here, \li{HTML} is a user-defined \emph{case type}, having cases for each HTML tag as well as some other structures, like text nodes and sequencing.
+Declarations  of some of these cases can be seen on lines 2-6 of Fig.
+\ref{f-htmltype} (note that TSL Wyvern also includes simple product types for convenience, written \li{T1 * T2}).
+We could again use Wyvern's general-purpose introductory form for case types, e.g.
+\li{HTML.BodyElement((attrs, child))} ({unlike in ML, in Wyvern we must explicitly qualify constructors with the case type they are part of when they are used}.
+This is largely to make our formal semantics simpler and for clarity of presentation.) But, as discussed above, using this syntax can be inconvenient and  cognitively demanding.
+Thus, we associate a TSL with \li{HTML} that provides a simplified notation for writing HTML, shown being used on lines 6-18 of Fig.
+\ref{f-example}.
+This literal body is layout-delimited, rather than delimited by explicit tokens as in Fig.
+\ref{f-delims}, and introduced by a form of \emph{forward reference}, written \li{~} (``tilde''), on the previous line.
+Because the forward reference occurs in a position where the expected type is \li{HTML}, the literal body is governed by that type's TSL.
+The forward reference will be replaced by the general-purpose term, of type \li{HTML}, generated by the TSL during typechecking.
+The use of layout imposes no syntactic constraints on the body, while using an inline form does (Fig.
+\ref{f-delims}).
+For HTML, this is quite useful, as all of the inline forms impose constrains that would cause conflict with some valid piece of HTML.
+\subsection{Implementing a TSL}
 \begin{figure}
@@
 \end{figure}
-On line 5 of Fig.
-\ref{f-example}, we see a call to a function \li{serve}, not shown, which has type \li{HTML -> Unit}.
-Here, \li{HTML} is a user-defined \emph{case type}, having cases for each HTML tag as well as some other structures, like text nodes and sequencing.
-Declarations  of some of these cases can be seen on lines 2-6 of Fig.
-\ref{f-htmltype} (note that TSL Wyvern also includes simple product types for convenience, written \li{T1 * T2}).
-We could again use Wyvern's general-purpose introductory form for case types, e.g.
-\li{HTML.BodyElement((attrs, child))} ({unlike in ML, in Wyvern we must explicitly qualify constructors with the case type they are part of when they are used}.
-This is largely to make our formal semantics simpler and for clarity of presentation.) But, as discussed above, using this syntax can be inconvenient and  cognitively demanding.
-Thus, we associate a TSL with \li{HTML} that provides a simplified notation for writing HTML, shown being used on lines 6-20 of Fig.
-\ref{f-example}.
-This literal body is layout-delimited, rather than delimited by explicit tokens as in Fig.
-\ref{f-delims}, and introduced by a form of \emph{forward reference}, written \li{~} (``tilde''), on the previous line.
-Because the forward reference occurs in a position where the expected type is \li{HTML}, the literal body is governed by that type's TSL.
-The forward reference will be replaced by the general-purpose term, of type \li{HTML}, generated by the TSL during typechecking.
-\subsection{Implementing a TSL}
-
 \begin{figure}[t]
@@
   metadata = new : HasTSL
-    val parser : Parser = (* quasiquotes *)
+    val parser : Parser = 
+      (* quasiquotes *)
 \end{lstlisting}
 \end{subfigure}
-\caption{Some of the types included in the Wyvern prelude.}
+\caption{Some of the types included in the Wyvern prelude. They are mutually defined.}
 \vspace{-10px}
@@
 \ref{f-htmltype}.
-A TSL is associated with a type, forming an \emph{active type}, using a more general mechanism for associating a value with a named type.
-A value associated with a named type is called its \emph{metadata}, and is introduced as shown on line 8 of Fig.
+A TSL is associated with a named type, forming an \emph{active type}, using a more general mechanism for associating a pure, static value with a named type, called its \emph{metadata}.
+Metadata is introduced as shown on line 8 of Fig.
 \ref{f-htmltype}.
+Type metadata, in this context, is comparable to class annotations in Java or attributes in C\#/F\# and internalizes the practice of writing metadata using comments, so that it can be checked by the language and accessed programmatically more easily.
+This can be used for a variety of purposes -- to associate documentation with a type, to mark types as being deprecated, and so on.
+
 For the purposes of this work, metadata values will always be of type \li{HasTSL}, an object type that declares a single field, \li{parser}, of type \li{Parser}.
-The \li{Parser} type is an object type declaring a single method, \li{parse}, that transforms a \li{TokenStream} extracted from a literal body to a Wyvern AST, which is a value of type \li{Exp}.
-This is a case type that encodes the abstract syntax of Wyvern expressions.
+The \li{Parser} type is an object type declaring a single method, \li{parse}, that transforms a \li{TokenStream} extracted from a literal body to a Wyvern AST.
+An AST is a value of type \li{Exp}, a case type that encodes the abstract syntax of Wyvern expressions.
 Fig.
-\ref{f-builtins} shows portions of the declarations of these types, which live in the Wyvern \emph{prelude} (a collection of ordinary Wyvern types that are automatically loaded before any other).
+\ref{f-builtins} shows portions of the declarations of these types, which live in the Wyvern \emph{prelude} (a collection of types that are automatically loaded before any other).
 
@@
 We formalize this hygiene mechanism by splitting the context in our static semantics.
-\item We provide several examples of TSLs throughout the paper, but to examine how broadly applicable the technique is, we conduct a simple corpus analysis, finding that string languages are used ubiquitously in existing Java code  (collaborative work with Darya Kurilova).
+\item To examine how broadly applicable the technique is, we conduct a simple corpus analysis, finding that string languages are used ubiquitously in existing Java code  (collaborative work with Darya Kurilova).
 \end{enumerate}
@@
 \subsection{Remaining Tasks and Timeline}
-The paper we have submitted to ECOOP 2014 has received quite positive reviews, so we anticipate that it will be accepted (notifications are sent on Mar.
-7th).
-If accepted, final revisions on the paper text are due on May 12th and so we will largely do these tasks after the OOPSLA deadline in late March.
-If not accepted, we will aim to resubmit to OOPSLA.
+This work has been accepted to ECOOP 2014, having received quite positive reviews.
+Final revisions on the paper text are due on May 12th and we will defer these tasks to after the OOPSLA deadline in late March.% If not accepted, we will aim to resubmit to OOPSLA.
 \begin{enumerate}
@@
 \item We must write down the full formal semantics (including rules that we omitted for concision in the paper draft) in a technical report, as well as provide more complete proofs of the metatheory.
-This might require slightly restricting the recursion in the rule for literals, so that induction is well-founded.
+This might require slightly restricting the recursion in the rule for literals, so that induction can be shown to be well-founded.
 \item Our current static semantics does not support mutually recursive type declarations, but this is necessary for our prelude to typecheck, so we will add support for this.
@@
 \begin{enumerate}
-\item We do not yet have a clean mechanism for preventing unintentional variable \emph{shadowing}, only unintentional variable \emph{capture}.
-It may be possible to prevent shadowing by preventing variables from leaking into interpolated Wyvern expressions (so that function application is the only way to pass data from a TSL to Wyvern code, as in the \li{Parser} TSL above).
+\item We have not yet formalized our mechanism for preventing unintentional variable \emph{shadowing}, only unintentional variable \emph{capture}.
+It should be possible to prevent shadowing by preventing variables from leaking into interpolated Wyvern expressions (so that function application is the only way to pass data from a TSL to Wyvern code, as in the \li{Parser} TSL above).
 \item In macro systems like Scheme's, free variables in generated ASTs refer to their bindings within the macro definition, rather than where they are inserted.
-To support this, we have introduced the \li{valAST} operator.
+To support this, we have introduced the \li{valAST} operator\todo{rename}.
 Using it is a bit verbose and awkward -- every free variable \li{x} in a quasiquote must instead be replaced with an interpolated term of the form \li{valAST(x)} for the resulting \li{Exp} to meaningful as the return value of the \li{parse} function.

### 2014-03-15 thesis-proposal/problem.tex
removed stuff

@@
 \section{Motivation}\label{motivation}
-Specifying and implementing a programming language together with its supporting tools (collectively, a \emph{programming system}) that is built upon sound theoretical foundations, helps users identify and fix errors as early as possible, supports a natural programming style, and performs well across diverse problem domains and hardware platforms remains a grand challenge in computing.
-In view of this goal, researchers and domain experts (collectively, \emph{providers}) continue to develop new special-purpose syntax, static and dynamic semantics, implementation strategies, optimizations, run-time systems and tools (collectively, \emph{features})  designed to address these challenges in increasingly diverse contexts.
-Ideally, a provider would be able to develop and distribute these kinds of new features orthogonally, as libraries, so that client developers could granularly choose those that best satisfy their needs.
-Unfortunately this is often infeasible because from the perspective of a library, the language's syntax and semantics are fully specified in advance, the compiler and run-time system are ``black box'' implementations of this fixed specification, and the other tools, like code editors and debuggers, operate according to fixed protocols.
-Providers of new system features must, as a result, take \emph{language-external approaches}, often by deriving a new programming system altogether.
-This approach has been encouraged, historically, by the availability of tools like compiler generators and language workbenches.
-We will argue that these approaches are technically problematic and that taking them has led to an unnecessary gap between research and practice.
-In their place, we will develop \emph{language-integrated extensibility mechanisms} that decentralize control over several core aspects of the programming system.
-By organizing new features around {types} and constraining them appropriately, we aim to show that these mechanisms can guarantee safety and non-interference of extensions while remaining highly expressive.
+%\begin{quote}\textit{The recent development of programming languages suggests that the simultaneous achievement of simplicity 
+%and generality in language design is a serious unsolved 
+%problem. }
+%- Reynolds, 1970\todo{citations}\end{quote}
+%%One might exp ``generality''
+Designing and implementing a simple but general-purpose programming language and its associated tools (collectively, a \emph{programming system}) has long been seen as a grand challenge in computing.
+Were an ideal system to emerge, we might expect that its experienced users would not often see the need to construct a new dialect of the language or its tools to implement a new high-level abstraction.
+A library-based implementations of an abstraction would be essentially as reasonable, natural and performant as a standalone implementation.
+By this metric, we are clearly far from our ideal.
+New language and tool dialects, originating both in research and in practice and spanning all major language paradigms, continue to be widely used as vehicles for new abstractions (examples of which we will discuss throughout this work), suggesting that not all useful  abstractions can be seen, when considered comprehensively, as mere modes of use of a contemporary ``general-purpose'' programming system.% The problem of achieving both simplicity and generality simultaneously with a single system remains unsolved.
+
+We observe that new system dialects are often motivated by the desire to introduce new syntax, strengthen the language's type system or provide specialized editor services, all aspects of existing systems that appear rather monolithic when considered from the perspective of a library provider.
+%, encouraged  historically  by the availability of tools like compiler generators and,  more recently, language workbenches \cite{workbenches} and DSL frameworks \cite{dsl}.
+Unfortunately, taking this approach makes it substantially more difficult for clients to import high-level abstractions orthogonally.
+It follows, then, that {language-integrated mechanisms} that allow library providers to extend these aspects of the system would decrease the need for new language and tool dialects, increasing the generality of the  system.
+However, such mechanisms must be introduced with care, because decentralizing control over these fundamental aspects of the system could destabilize it by weakening its metatheory, permitting ambiguities and conflicts and obscuring the semantics of a piece of code.
+In this thesis, we aim to show that by associating language extensions with user-defined types (forming what we call \emph{active types}) and structuring the underlying extension mechanism to enforce critical  abstraction barriers, we can give library providers the ability to express powerful extensions to the syntax, the type system and even to editor services from within libraries, while avoiding these kinds of stability issues.
+%We call user-defined types that introduce new features into the system in this way \emph{active types}.
+% also highly expressive.
 %But taking a \emph{language-internal approach} to implementing a feature is the most practical.
@@
 \subsection{Motivating Example: Regular Expressions}\label{regex}
-To make the problem we aim to address concrete, we begin with a simple example that we will return to throughout this work.
-\emph{Regular expressions} are a widely-used abstraction for finding patterns in semi-structured strings (e.g.
+To make the problems we aim to address more concrete, we begin with a simple example that we will return to throughout this work.
+\emph{Regular expressions} are commonly used to express patterns in strings (e.g.
 DNA sequences) \cite{Thompson:1968:PTR:363347.363387}.
-If a programming system wished to fully support regular expressions, it might simultaneously provide features like these:
+If a programming system provided full support for this abstraction, it might  provide, or enable the provision of, features like:
 
 \begin{enumerate}
-\item \textbf{Syntax for pattern literals} so that the cognitive load of reading and writing patterns is low, patterns are parsed once at compile-time and malformed patterns result in intelligible {compile-time} errors.
-\item A \textbf{static semantics} that ensures that key invariants related to regular expressions are maintained:
+\item \textbf{Syntax for pattern literals}.
+An ideal syntax would permit us to express patterns concisely and clearly.
+For example, The BisI restriction enzyme cuts DNA whenever it sees the pattern \texttt{GC\textit{N}GC}, where \texttt{\textit{N}} represents any base.
+We would want to express it, perhaps, as follows (using curly braces to splice one pattern into another):
+\begin{lstlisting}[numbers=none]
+let N : Pattern = <SURLA|T|G|CEURL>
+let BisI : Pattern = <SURLGC{EURLNSURL}GCEURL>\end{lstlisting}
+%The cognitive load of reading and writing patterns is low and patterns are parsed once at compile-time. 
+Malformed patterns would result in intelligible {compile-time} errors.
+\item A \textbf{type system} that ensures that key invariants related to regular expressions are statically maintained:
 	\begin{enumerate}
-	\item only other patterns and properly escaped strings are interpolated into a pattern, to avoid splicing errors and injection attacks \cite{owasp2013, Bravenboer:2007:PIA:1289971.1289975}
-	\item out-of-bounds backreferences are not used \cite{spishak2012type}
-	\item string processing operations do not lead to a string that is malformed, when well-formedness can be captured by a regular expression \cite{fulton-thesis}
+	\item only other patterns and properly escaped strings are spliced into a pattern, to avoid splicing errors and injection attacks \cite{owasp2013, Bravenboer:2007:PIA:1289971.1289975}
+	\item out-of-bounds backreferences to a captured group are not used \cite{spishak2012type}
+	\item string processing operations do not lead to a string that is malformed, when well-formedness can be captured as membership in a regular language \cite{fulton-thesis}
 	\end{enumerate}
-When an error is found, an intelligible error message is provided.
+%When an error is found, an intelligible error message is provided.
 %\item An \textbf{implementation} that partially or fully compiles known regular expressions into the efficient internal representation that will be used by the regular expression matching engine (e.g., a finite automata \cite{Thompson:1968:PTR:363347.363387}) ahead of time.
@@
 %By performing this step ahead-of-time, these dangers can be avoided.
-\item \textbf{Editor services} that allow clients to interactively test regular expression patterns against test strings, refer to documentation or search for common patterns (e.g.
-\cite{IntelliJRegexp}).
+\item \textbf{Editor services} that allow clients to refer to documentation, interactively test patterns against test strings and search for common patterns (e.g. \cite{IntelliJRegexp})
 \end{enumerate}
@@
 %In a conventional \emph{monolithic} programming system, support for each of these features would need to be built into the language and tools. 
-No system today builds in support for all of the features enumerated above in their strongest form.
-Instead, libraries generally provide support for regular expressions by leveraging  general-purpose abstraction mechanisms.
-Unfortunately, it is impossible to fully define the syntax and the specialized static semantics described in the references above in terms of general-purpose notations and abstractions.
+No system today builds in support for all of the features enumerated above in their strongest form, so library providers must provide support for regular expressions by leveraging  general-purpose constructs.
+Unfortunately, it is impossible to completely define the syntax and the specialized static semantics described in the references above in terms of general-purpose notations and abstractions.
 Library providers thus need to compromise.
-The most common strategy is to ask clients to enter patterns as strings deferring their parsing, typechecking, compilation and use to run-time.
-This proves only a weak approximation to feature 1 (due to clashes between string escape sequences and regular expression syntax).
-None of the static guarantees can be provided in this way, leading to run-time exceptions (even in well-tested code, as shown in \cite{spishak2012type}), and logic errors that are not caught even at run-time, some of which can lead to security vulnerabilities (due to injection attacks, for example \cite{owasp2013}).
-It also introduces performance overhead (due to run-time parsing and compilation of patterns, and potentially redundant run-time string checks).
-Requiring that clients instead introduce patterns directly using general-purpose constructs like object or inductive datatype constructors or operations over an abstract type rather than via strings can only provide feature 2a while eliminating even the weak approximation to feature 1 that the use of string literals provides.
 
-None of these approaches address the issue of tool support.
-Regular expressions are not trivial to work with, so tool support can be quite helpful, even for experts, and a number of tools are available online.
-Unfortunately, tools that must be discovered independently and accessed externally are used infrequently \cite{Murphy-Hill:2011:PIE:1958824.1958888} and are less usable than editor-integrated tools \cite{Campbell:2008:DRT:1636642.1636651, Omar:2012:ACC:2337223.2337324}, leading to lower productivity.
+The most common strategy is to ask clients to enter patterns as strings, deferring their parsing, typechecking, compilation and use all to run-time.
+This provides only a weak approximation to feature 1 (due to clashes between string escape sequences and regular expression syntax).
+None of the static guarantees can be provided in this way, leading to run-time exceptions (even in well-tested code \cite{spishak2012type}), and logic errors that cannot easily be caught even at run-time, some of which can lead to security vulnerabilities (due to injection attacks \cite{owasp2013}).
+It also introduces performance overhead due to run-time parsing and compilation of patterns, and, in the absence of feature 2c,   redundant run-time string checks.
+Requiring that clients instead introduce patterns directly using general-purpose constructs like object or datatype constructors or operations over an abstract type rather than via strings provides only feature 2a at the expense of the weak approximation to feature 1 that the use of string literals provided.
 
+None of these approaches address the issue of tool support (feature 3).
+Regular expressions are not trivial to work with, and we have found that tool support can be quite helpful, even for professional programmers \cite{Omar:2012:ACC:2337223.2337324}.
+A number of tools are available online that provide helpful services like syntax highlighting for patterns and pattern extraction from examples (e.g.
+\cite{_txt2re:_????}).
+Unfortunately, tools that must be accessed externally are difficult to discover and are thus used infrequently \cite{Murphy-Hill:2011:PIE:1958824.1958888,Campbell:2008:DRT:1636642.1636651,Omar:2012:ACC:2337223.2337324}.
+They are also less usable than editor-integrated tools (e.g.
+\cite{IntelliJRegexp}) because they require the programmer to switch contexts  and cannot make use of the editor context.
+Working with high-level abstractions in a simple text editor is thus awkward and error-prone, just like attempting to work with high-level abstractions in a strictly low-level language.
+For these reasons, we take the position that the editor is as integral to a good programming system as the language itself (though not the position that it be a part of a full integrated development environment).
 \subsection{Language-External Approaches}\label{external-approaches}
@@
 (b) When taking a language-integrated approach, there is one extensible host language and the compile-time and edit-time logic governing new constructs is expressed within ``active'' libraries.
-If the extension mechanism guarantees that extensions cannot interfere with one another or the base system, these problems are avoided.}
+If the extension mechanism guarantees that extensions cannot cause instabilities, the problems of orthogonality and client compatibility are avoided.}
 \label{approaches}
@@
 
-When the syntax and semantics of a system must be extended to fully realize a new feature, as in the example above and others we will give throughout this work, providers typically take a \emph{language-external approach}, either by developing a new or derivative programming system (supported by \emph{language workbenches} \cite{erdweg2013state},  \emph{DSL frameworks} \cite{fowler2010domain} or \emph{compiler generators} \cite{brooker1963compiler}), or by extending an existing artifact, such as an extension mechanism for a {particular} compiler\footnote{Compilers that modify, or allow modification of, the semantics of their base language, rather than simply permitting semantics-preserving optimizations, should be considered a pernicious means for creating new languages.
-Many ``practical'' compilers, including \texttt{gcc}, \texttt{GHC} and \texttt{SML/NJ}, are guilty of this, meaning that some programs that seem to be written in C, Haskell or Standard ML are actually written in tool-specific derivatives of these languages.
-Language-integrated mechanisms do not lead to such fragmentation.}, editor or other tool.
+When the syntax and semantics of a system must be extended to fully realize a new feature providers typically take a \emph{language-external approach}, either by developing a new system dialect (supported by \emph{compiler generators} \cite{brooker1963compiler}, \emph{language workbenches} \cite{erdweg2013state}, \emph{DSL frameworks} \cite{fowler2010domain}, or simply be forking an existing codebase), or by using an extension mechanism for a {particular} compiler\footnote{Compilers that modify, or allow modification of, the semantics of their base language, rather than simply permitting semantics-preserving optimizations, should be considered a pernicious means for creating new languages.
+That is, some programs that purport to be written in C, Haskell or Standard ML are actually written in compiler-specific dialects of these languages.}, editor or other tool.
 For example, a researcher interested in providing regular expression related features (let us refer to these collectively as \texttt{R}) might design a new system with built-in support for these, perhaps basing it on an existing system containing some general-purpose features (\texttt{G}).
 A different researcher developing a new language-integrated parallel programming abstraction (\texttt{P}) might  take the same approach.
-A third researcher, developing an alternative parallel programming abstraction (\texttt{Q}) might again do the same.
-This results in a collection of distinct systems as diagrammed in Figure \ref{approaches}a.
-Unfortunately, when providers of new features take language-external approaches like this, it causes  problems for clients related to \textbf{orthogonality} and \textbf{client compatibility}.
-%This latter method couples the semantics of the feature to the implementation details of a particular tool.
+A third researcher, developing a type system for reasoning about units of measure (\texttt{Q}) might again do the same.
+This results in a collection of distinct systems, as diagrammed in Figure \ref{approaches}a.
+This might be entirely sufficient if developing proofs of concept is the only goal, but when providers of new features take language-external approaches like this, it causes  problems for clients (including other researchers) related to \textbf{orthogonality} and \textbf{client compatibility}.
+This has limited the broad adoption of these kinds of innovations.%This latter method couples the semantics of the feature to the implementation details of a particular tool.
 Because the use of one implementation entails a different semantics for the feature than another, the extended tool acts, \emph{de facto}, as a distinct system for our purposes.
@@
 That is, one must either use the system containing features \texttt{G+R}, \texttt{G+P} or \texttt{G+Q}.
-There is no system containing \texttt{G}, \texttt{R}, \texttt{P} and \texttt{Q} in other combinations, and merging the systems containing each separately can be non-trivial because there is no requirement that providers of new features use a common mechanism.
-Even in cases where a common mechanism has been used, there are serious interference and safety issues, as we will discuss at length throughout this thesis.
+There is no system containing \texttt{G}, \texttt{R}, \texttt{P} and \texttt{Q} in other combinations, and merging the systems containing each separately can be non-trivial because, even in cases where a common mechanism has been used, there are serious interference and safety issues, as we will discuss at length throughout this thesis.
 
-Recent evidence indicates that this is one of the major barriers preventing research from being driven into practice: developers prefer language-integrated parallel programming abstractions with stronger safety guarantees and more natural syntax to library-based abstractions if all else is equal \cite{cave2010comparing}, but library-based implementations are more widely adopted because ``parallel programming languages'' privilege only a few chosen  abstractions at the language level.
+Recent evidence indicates that this is one of the major barriers preventing research from being driven into practice.
+For example, developers prefer language-integrated parallel programming abstractions with stronger safety guarantees and more natural syntax to library-based abstractions when all else is equal \cite{cave2010comparing}, but library-based implementations are more widely adopted because ``parallel programming languages'' privilege only a few chosen  abstractions at the language level.
 This is problematic because different parallel programming abstractions are seen as more appropriate in different situations \cite{Tasharofi:2013rc}.
-Moreover,  parallel programming support is rarely the only concern relevant to client developers outside of a classroom setting.
-Regular expression support, for example, may be simultaneously desirable for processing large amounts of textual data in parallel, but using these features together in the same compilation unit would be difficult or impossible.
-%Similarly, a language and tools designed primarily to support regular expressions might make an interesting research project, but it would not be a suitable tool for writing large applications with more varied needs.
+Moreover,  parallel programming support is rarely the only concern relevant to clients outside of a classroom setting.
+Support for regular expressions, for example, would be simultaneously desirable for processing large amounts of genomic data in parallel, but using these features together in the same compilation unit would be difficult or impossible.
+Indeed, switching to a ``parallel programming language'' would likely make it \emph{more} difficult to use regular expressions, as these are likely to be less well-developed in a specialized language than in an established general-purpose language.
+The intuition was perhaps most succinctly expressed by a participant in a recent study by Basili et al.
+\cite{basili2008understanding}:  ``I hate MPI, I hate C++.
+[But] if I had to choose again, I would probably choose the same.'' %Similarly, a language and tools designed primarily to support regular expressions might make an interesting research project, but it would not be a suitable tool for writing large applications with more varied needs.
 
@@
 
-\paragraph{Client Compatibility} Even in cases where for each component of a software system there is a programming system that is completely satisfactory in isolation, there remain problems at the interface between components.
-An interface that exposes the specialized constructs particular to one language (e.g.
-futures in a parallel programming language) cannot necessarily be safely and naturally consumed from another language (e.g.
-a general-purpose language).
-Tool support is also lost when calling into a different language.
-We call this fundamental issue the \emph{client compatibility problem}: code written by clients of a certain collection of features cannot always interface with code written by clients of a different collection  in a safe, performant and natural manner.
+\paragraph{Client Compatibility} Even in cases where, for each component of a software system, there is a programming system considered entirely satisfactory by its developers, there remain problems at the interface between components.
+An interface that exposes externally a specialized construct particular to one language (e.g.
+a function that requires a quantity having a particular unit of measure) cannot necessarily be safely and naturally consumed from another language (e.g.
+a parallel programming language).
+Tool support is also lost when calling into different languages.
+We call this the \emph{client compatibility problem}: code written by clients of a certain collection of features cannot always interface with code written by clients of a different collection  in a safe, performant and natural manner.
 
-One strategy often taken by proponents of a \emph{language-oriented approach} to software development \cite{journals/stp/Ward94} to partially address the client compatibility problem is to  target an established intermediate language, such as the Java Virtual Machine (JVM) bytecode, and use its constructs as a common language for communication between components written in different languages.
+One strategy often taken by proponents of a {language-oriented approach} to abstraction development \cite{journals/stp/Ward94} to partially address the client compatibility problem is to  target an established intermediate language and use its constructs as a common language for communication between components written in different languages.
 Scala \cite{200464/IC} and F\# \cite{pickering2007foundations} are examples of prominent general-purpose languages that have taken this approach, and most DSL frameworks also rely on this strategy.
 As indicated in Figure \ref{approaches}a, this only enables client compatibility in one direction.
-Calling into the common language becomes straightforward and safe, but calling in the other direction, or between the languages sharing the common target, does not.
+Calling into the common language becomes straightforward and safe, but calling in the other direction, or between the languages sharing the common target, does not, unless these languages are only trivially different from the intermediate language.
 
-This approach can work well when new languages consist of constructs that can also be expressed safely and almost as naturally in the common language.
-But many of the most innovative constructs found in modern languages (often, those that justify their creation) are difficult to define in terms of existing constructs in ways that guarantee all necessary invariants are statically maintained and that do not require large amounts boilerplate code and run-time overhead.
-As a trivial example with significant practical implications, F\#'s type system does not admit \verb|null| as a value for any type defined within F\# code, but maintaining this important invariant still requires run-time checks because the typing rules of F\# do not apply when F\# data structures are constructed by other languages on the Common Language Infrastructure (CLI) like C\# or SML.NET.
+%This approach only works well when new languages consist of constructs that can also be expressed safely and almost as naturally in the common language.
+%But many of the most innovative constructs found in modern languages (often, those that justify their creation) are difficult to define in terms of existing constructs in ways that guarantee all necessary invariants are statically maintained and that do not require large amounts boilerplate code and run-time overhead.
+As a simple example with significant contemporary implications, F\#'s type system does not admit \verb|null| as a value for a type introduced by F\#, but maintaining this sensible internal invariant still requires dynamic  checks because the stricter typing rules of F\# do not apply when F\# data structures are constructed by other languages on the Common Language Infrastructure (CLI) like C\# or SML.NET.
 This is not an issue exclusive to intermediate languages that make regrettable choices regarding \verb|null|, however.
-The F\# type system also includes support for checking that units of measure are used correctly \cite{syme2012expert, kennedy1994dimension}, but this more specialized invariant is left unchecked at language boundaries.
-No alternative  general-purpose intermediate language would change this situation.
-%Even exposing datatypes  and tuples at component boundaries, which are used ubiquitously and can be safely consumed from other languages, is not recommended when a component might be used from another language, according to drafted F\# component design guidelines \cite{fsharpguidelines}, because these are awkward to consume from other languages without the convenient primitive operations (e.g.
+The F\# type system also includes support for checking that units of measure are used correctly \cite{syme2012expert, kennedy1994dimension}, but this more specialized static invariant is left entirely unchecked at language boundaries.
+Exposing functions that operate over datatypes, tuples and values having units of measure is not recommended when a component ``might be used'' from another language, according to drafted F\# component design guidelines \cite{fsharpguidelines}, because it is awkward to construct and consume these from other languages without the convenient primitive operations (e.g.
 pattern matching) and syntax that F\# includes.
-SML.NET prohibits exposing such types at component boundaries altogether.
+SML.NET prohibits exposing such types at component boundaries altogether (and cannot naturally consume F\# data structures, despite having a rather similar syntax and semantics in many ways).
 %In Scala, traits that have default method implementations are difficult to implement from Java or other JVM languages and the workaround can break if the trait is modified \cite{scalatraitinterop}. 
@@
 so that language-external approaches are less frequently necessary.
-More specifically, we will show how control over aspects of the \textbf{syntax}, \textbf{static and dynamic semantics} and \textbf{editor services} can be delegated to user-defined logic distributed in {libraries}, as illustrated in Figure \ref{approaches}b.
+More specifically, we will show how control over aspects of the \textbf{syntax}, \textbf{type system} and \textbf{editor services} can be delegated to user-defined logic distributed in {libraries}, as illustrated in Figure \ref{approaches}b.
 Such libraries have been called \emph{active libraries}  \cite{activelibraries} because, rather than being passive clients of features already available in the system, they contain logic invoked by the system during development or compilation to provide new features.
@@
 Furthermore, even if extensions can be shown not to cause such problems in isolation, there may still be conflicts between extensions that could weaken their semantics, leading to subtle problems that only appear when two extensions are used together.
-For example, if two active libraries introduce the same syntactic form but back it with differing (but individually valid) semantics, the issue would only manifest itself when both libraries were imported somewhere within the same scope.
-These kinds of safety issues have plagued previous attempts to design language-integrated extensibility mechanisms.
+As a simple example, if two active libraries introduce the same syntactic form but back it with differing (but individually valid) semantics, the issue would only manifest itself when both libraries were imported within the same scope.
+Resolving these kinds of ambiguities requires significantly more expertise with parser technology than using the syntax itself does.
+These and more serious safety and ambiguity issues have plagued previous attempts to design language-integrated extensibility mechanisms.
 We will briefly review some of these attempts below.% To prevent them, our mechanisms will organize  extension logic around types to guarantee that extensions are both safe in isolation and also safely composable in any combination.

### 2014-03-17 sciunit-icse14/sciunit-icse14-final/sciunit-icse14.tex
Updated ICSE14 and proposal

--- 2014-03-15 sciunit-icse14/sciunit-icse14.tex
@@
 The core SciUnit framework has been fully developed and is available at http://sciunit.scidash.org/. SciDash is under active development as of this writing.
-\vspace{5px}
+\vspace{2px}
  %In particular, we consider modeling competitions in science as a form of test-driven validation.
@@
 \section{ACKNOWLEDGEMENTS}
-We thank Sharon Crook, Shreejoy Tripathy and Padraig Gleeson for their many helpful discussions, and anonymous referees and several friends for their careful readings.  
-RG was supported in part by NIMH grant R01MH081905. 
-The content is solely the responsibility of the authors and does not necessarily represent the official views of the NIH.
-\vspace{5px}
+\vspace{-1px}
+We thank Sharon Crook, Shreejoy Tripathy and Padraig Gleeson for their many helpful discussions, and anonymous referees and friends for their careful readings.  
+RG was supported in part by NIMH grant R01MH081905. CO and JA were supported in part by NSF grant 1116907. 
+The content is solely the responsibility of the authors and does not necessarily represent the official views of the NIH or NSF.
+\vspace{2px}
 % conference papers do not normally have an appendix
@@
 %\renewcommand\bibsection{\subsection{\refname}}
-\renewcommand\refname{New Title}
 \bibliographystyle{abbrv}

### 2014-03-17 thesis-proposal/parsing.tex
Updated ICSE14 and proposal

@@
 \section{Type-Specific Languages}\label{aparsing}
-General-purpose constructs like those built into modern languages are, semantically, quite powerful.
-By using a general-purpose construct to define a data structure, one immediately benefits from a body of useful operators, established reasoning principles, well-optimized implementations and tool support.
-For example, lists can be encoded using a more general mechanism for defining polymorphic recursive sum types.
-Intuitively, a list can either be empty, or be broken down into a \emph{head} element and a \emph{tail}, another list.
-In an ML-like language, the polymorphic list type would be defined:
+General-purpose constructs are, semantically, quite flexible. 
+%By using a general-purpose construct to define a data structure, one immediately benefits from a body of useful operators, established reasoning principles, well-optimized implementations and tool support.
+For example, lists can be defined using a more general mechanism for defining polymorphic recursive sum types by observing that a list can either be empty, or be broken down into a \emph{head} element and a \emph{tail}, another list.
+In an ML-like language, this would be written:
 \begin{lstlisting}[numbers=none]
@@
 \end{lstlisting}
-By defining lists in this way, we can immediately reason about list values by structural induction and examine them by pattern matching.
+By defining abstractions in terms of existing general purpose constructs, we immediately know how to reason about them (here, by structural induction) and examine them (by pattern matching).
+They are already well optimized by compilers and benefit from editor support as well.
 %The programmer only needs to provide an encoding of the structure of lists; the semantics and implementation are filled in by the general-purpose abstraction mechanism.
-While this is quite useful, the associated general-purpose notation is not. For example, few would claim that writing a list of numbers as a sequence of \li{Cons} cells is convenient:
+While this is all quite useful, the associated general-purpose syntax is often not. For example, few would claim that writing a list of numbers as a sequence of \li{Cons} cells is convenient:
 \begin{lstlisting}[numbers=none]
@@
 Using terminology from Green's cognitive dimensions of notations \cite{green1996usability}, it is more \emph{terse}, \emph{visible} and \emph{maps more closely} to the intuitive notion of a list.
-Stoy, in writing about the value of good notation, says \cite{stoybook}:
+Stoy, in discussing the value of good notation, write \cite{stoy1977denotational}:
 \begin{quote}A good notation thus conceals much of the inner workings behind suitable abbreviations, while allowing us to consider it in more detail if we require: matrix and tensor notations provide further good examples of this.
@@
 This is clearly more cognitively demanding, both when writing the regular expression and when reading it.
-Among the most common strategies in these situations, for users of any kind of language, is again to simply use a string representation that is parsed at run-time.
+Among the most common strategies in these situations, for users of any kind of language, is again to simply use a string representation that is parsed at run-time:
 \begin{lstlisting}
@@
 
-So although languages that allow users to introduce new syntax from within libraries appear to hold promise for the reasons described above, providing this form of extensibility is non-trivial because there is no longer a central designer making decisions about such ambiguities.
+So although languages that allow providers to introduce new syntax from within libraries appear to hold promise for the reasons described above, enabling this form of extensibility is non-trivial because there is no longer a central designer making decisions about such ambiguities.
 In most existing related work, the burden of resolving ambiguities falls to clients who have the misfortune of importing conflicting extensions.
@@
 two regular expression engines) can all cause problems that may be difficult to anticipate.
+Techniques that limit the kinds of syntax extensions that can be expressed to guarantee ambiguities cannot occur simply cannot express these kinds of examples as-is (e.g.
+\cite{conf/pldi/SchwerdfegerW09}).
 
@@
 This strategy avoids the problem of conflicting syntax, because neither the base language nor TSLs are ever extended directly.
-It also permits semantic flexibility -- the meaning of a form like \verb|{ }| can differ depending  on its type, so it is safe to use it for empty sets, maps and other data structures, like JSON literals.
-This frees these common notations from being tied to the variant of a  data structure built into a language's standard library, which sometimes does not provide the exact semantics that a programmer needs (for example, Python dictionaries do not preserve key insertion order).
+It also permits semantic flexibility -- the meaning of a form like \verb|{ }| can differ depending  on its type, so it is safe to use it for empty sets, maps and other data structures, like JSON literals,  freeing these common notations from being tied to the variant of a  data structure built into a language's standard library, which sometimes does not provide the exact semantics that a programmer needs (for example, Python dictionaries do not preserve key insertion order).
 \begin{figure}[t]
@@
 new : SearchServer
-  def resultsFor(searchQuery : String, page : Nat) : Unit =
+  def resultsFor(searchQuery, page) =
     serve(~) (* serve : HTML -> Unit *)
-SHTML      :html
-        :head
-          :title Search Results
-          :style EHTML~
+SHTML      >html
+        >head
+          >title Search Results
+          >style EHTML~
 SCSS            body { background-image: url(%ECSSbgImageSCSS%) }
-            #search { background-color: %ECSS`SCOLOR#aabbccECOLOR`SCSS% }
-ECSSSHTML        :body
-          :h1 Results for :{EHTMLsearchQuerySHTML}
-          :div[id="search"]
-            Search again: :{EHTMLSearchBox($\texttt{"}$SSTRGo!ESTR$\texttt{"}$)SHTML}
-          :wyvEHTML (* fmt_results : DB * SQLQuery * Nat * Nat -> HTML *)
+            #search { background-color: %ECSSdarken(`SCOLOR#aabbccECOLOR`, SPCT10pctEPCT)SCSS% }
+ECSSSHTML        >body
+          >h1 Results for < EHTMLHTML.Text(searchQuery)SHTML
+          >div[id="search"]
+            Search again: < EHTMLSearchBox($\texttt{"}$SSTRGo!ESTR$\texttt{"}$)SHTML
+          <EHTML (* fmt_results : DB * SQLQuery * Nat * Nat -> HTML *)
              fmt_results(db, ~, SNAT10ENDNAT, page)
@@
 To allow us to focus on the essence of our proposal, the variant of Wyvern we will describe in this thesis is simpler than the variant previously described: it is purely functional (there are no effects other than non-termination) and it does not enforce a uniform access principle for objects (fields can be accessed directly).
-Objects can thus be thought of as functional records equipped with simple methods (functions that are automatically given a self-reference) for convenience.
+Objects can thus be thought of as recursive labeled products with support for with simple methods (functions that are automatically given a self-reference) for convenience.
 We also add recursive labeled sum types, which we call \emph{case types}, that are quite similar to datatypes in ML.
 One can refer to the version of the language described in this thesis as \emph{TSL Wyvern}.
+TSL Wyvern has a layout-sensitive syntax, for reasons we will discuss.
 
@@
   metadata = new : HasTSL
-    val parser : Parser = ~
-SGRM      start -> ':body'= start>
-        EGRMfn child:Exp => `SQTHTML.BodyElement(([], %EQTchildSQT%))EQT`
-SGRM      start -> ':style'= EXP>
-        EGRMfn e:Exp => `SQTHTML.StyleElement(([], %EQTeSQT%))EQT`
-SGRM      start -> '{'= EXP '}'=EGRM
-        fn e:Exp => `SQT%EQTeSQT% : HTMLEQT`
+    val parser = ~
+SGRM      start <- '>body'= start>
+        EGRMfn child => `SQTHTML.BodyElement(([], %EQTchildSQT%))EQT`
+SGRM      start <- '>style'= EXP>
+        EGRMfn e => `SQTHTML.StyleElement(([], %EQTeSQT%))EQT`
+SGRM      start <- '<'= EXP>EGRM 
+        fn e => `SQT%EQTeSQT% : HTMLEQT`
 \end{lstlisting}
@@
 \begin{figure}[t]
-\begin{subfigure}[t]{.53\textwidth}
+\begin{subfigure}[t]{.58\textwidth}
 \begin{lstlisting}
@@
 \end{subfigure}
-\begin{subfigure}[t]{.55\textwidth}
-\begin{lstlisting}[linewidth=.5\textwidth]
+\begin{subfigure}[t]{.42\textwidth}
+\begin{lstlisting}[linewidth=.42\textwidth,firstnumber=14]
 casetype Exp 
@@
   metadata = new : HasTSL
-    val parser : Parser = 
-      (* quasiquotes *)
+    val parser : Parser = (* exp
+      quasiquotes *)
+...
 \end{lstlisting}
@@
 Each production defines a sequence of terminals (e.g.
-\li{':body'}) and non-terminals (e.g.
+\li{'>body'}) and non-terminals (e.g.
 \li{start}, or one of the built-in non-terminals \li{ID}, \li{EXP} or \li{TYPE}, representing Wyvern identifiers, expressions and types, respectively).
@@
 \item In macro systems like Scheme's, free variables in generated ASTs refer to their bindings within the macro definition, rather than where they are inserted.
-To support this, we have introduced the \li{valAST} operator\todo{rename}.
-Using it is a bit verbose and awkward -- every free variable \li{x} in a quasiquote must instead be replaced with an interpolated term of the form \li{valAST(x)} for the resulting \li{Exp} to meaningful as the return value of the \li{parse} function.
-We believe we can insert \li{valAST} automatically from within the definition of the TSL for \li{Exp} using single variable (rather than whole expression) interpolation.
+To support this, we have introduced the \li{toast} operator.
+Using it is currently explicit.
+We believe we can insert \li{toast} automatically using single variable (rather than whole expression) interpolation.
 \end{enumerate}

### 2014-03-17 thesis-proposal/problem.tex
Updated ICSE14 and proposal

@@
 %%One might exp ``generality''
-Designing and implementing a simple but general-purpose programming language and its associated tools (collectively, a \emph{programming system}) has long been seen as a grand challenge in computing.
-Were an ideal system to emerge, we might expect that its experienced users would not often see the need to construct a new dialect of the language or its tools to implement a new high-level abstraction.
-A library-based implementations of an abstraction would be essentially as reasonable, natural and performant as a standalone implementation.
+Specifying and implementing a simple but general-purpose programming language and its associated tools (collectively, a \emph{programming system}) has long been seen as a grand challenge in computing.
+Were an ideal system to emerge, we might expect that providing a new high-level abstraction would not require specifying a new dialect of the language or its tools, because a library-based embedding of the abstraction would be essentially as reasonable, natural and performant.
 By this metric, we are clearly far from our ideal.
-New language and tool dialects, originating both in research and in practice and spanning all major language paradigms, continue to be widely used as vehicles for new abstractions (examples of which we will discuss throughout this work), suggesting that not all useful  abstractions can be seen, when considered comprehensively, as mere modes of use of a contemporary ``general-purpose'' programming system.% The problem of achieving both simplicity and generality simultaneously with a single system remains unsolved.
+New language and tool dialects, originating both in research and in practice and spanning all major language lineages, continue to be widely used as vehicles for new abstractions (examples of which we will discuss throughout this work).
+This suggests that not all useful  abstractions can be seen, when considered comprehensively, as mere modes of use of a contemporary ``general-purpose'' programming system.% The problem of achieving both simplicity and generality simultaneously with a single system remains unsolved.
 
-We observe that new system dialects are often motivated by the desire to introduce new syntax, strengthen the language's type system or provide specialized editor services, all aspects of existing systems that appear rather monolithic when considered from the perspective of a library provider.
+We observe that new dialects are often motivated by the desire to introduce new syntax, strengthen the language's type system or provide specialized editor services, all aspects of existing systems that appear rather monolithic when considered from the perspective of a library provider.
 %, encouraged  historically  by the availability of tools like compiler generators and,  more recently, language workbenches \cite{workbenches} and DSL frameworks \cite{dsl}.
 Unfortunately, taking this approach makes it substantially more difficult for clients to import high-level abstractions orthogonally.
-It follows, then, that {language-integrated mechanisms} that allow library providers to extend these aspects of the system would decrease the need for new language and tool dialects, increasing the generality of the  system.
-However, such mechanisms must be introduced with care, because decentralizing control over these fundamental aspects of the system could destabilize it by weakening its metatheory, permitting ambiguities and conflicts and obscuring the semantics of a piece of code.
-In this thesis, we aim to show that by associating language extensions with user-defined types (forming what we call \emph{active types}) and structuring the underlying extension mechanism to enforce critical  abstraction barriers, we can give library providers the ability to express powerful extensions to the syntax, the type system and even to editor services from within libraries, while avoiding these kinds of stability issues.
+It follows, then, that {language-integrated mechanisms} that allow library providers to extend these features of the system would increase its generality by decreasing the need for new language and tool dialects.
+However, such mechanisms must be introduced with care, because decentralizing control over such fundamental aspects of the system could destabilize it by weakening its metatheory and permitting ambiguities and conflicts (collectively, \emph{safety issues}).
+It could also make it more difficult to reason about the meaning of code.
+In this thesis, we aim to show that by adopting the natural convention of organizing language extensions around user-defined types (forming what we call \emph{active types}) and designing an underlying type system that enforces critical  abstraction barriers between extensions, we can give abstraction providers the ability to express from within libraries powerful, orthogonal extensions to the syntax and type system, and define powerful editor services, without the possibility of the kinds of safety issues that have limited prior approaches.
 %We call user-defined types that introduce new features into the system in this way \emph{active types}.
@@
 DNA sequences) \cite{Thompson:1968:PTR:363347.363387}.
-If a programming system provided full support for this abstraction, it might  provide, or enable the provision of, features like:
+If a programming system provided full support for this abstraction, it might  provide features like:
 
@@
 %By performing this step ahead-of-time, these dangers can be avoided.
-\item \textbf{Editor services} that allow clients to refer to documentation, interactively test patterns against test strings and search for common patterns (e.g. \cite{IntelliJRegexp})
+\item \textbf{Editor services} to support syntax highlighting of patterns and to allow clients to refer to documentation and interactively test patterns against examples strings (e.g.
+\cite{IntelliJRegexp}) or even extract patterns from examples interactively (e.g.
+as in \cite{_txt2re:_????}).
 \end{enumerate}
@@
 No system today builds in support for all of the features enumerated above in their strongest form, so library providers must provide support for regular expressions by leveraging  general-purpose constructs.
-Unfortunately, it is impossible to completely define the syntax and the specialized static semantics described in the references above in terms of general-purpose notations and abstractions.
-Library providers thus need to compromise.
+Unfortunately, it is impossible to completely define the syntax and the specialized static semantics described in the references above in terms of general-purpose notations and abstractions, so library providers need to compromise.
 
@@
 Unfortunately, tools that must be accessed externally are difficult to discover and are thus used infrequently \cite{Murphy-Hill:2011:PIE:1958824.1958888,Campbell:2008:DRT:1636642.1636651,Omar:2012:ACC:2337223.2337324}.
-They are also less usable than editor-integrated tools (e.g.
-\cite{IntelliJRegexp}) because they require the programmer to switch contexts  and cannot make use of the editor context.
-Working with high-level abstractions in a simple text editor is thus awkward and error-prone, just like attempting to work with high-level abstractions in a strictly low-level language.
-For these reasons, we take the position that the editor is as integral to a good programming system as the language itself (though not the position that it be a part of a full integrated development environment).
+We found in experiments that they are also less usable than editor-integrated tools %(e.g.
+\cite{IntelliJRegexp})
+because they require the programmer to switch contexts  and cannot make use of the code context \cite{Omar:2012:ACC:2337223.2337324}.
+Working with high-level abstractions in a simple editor is thus awkward and error-prone, just like attempting to work with high-level abstractions in a low-level language.
+For these reasons,   we consider editor behavior as within the purview of an abstraction specification.
 \subsection{Language-External Approaches}\label{external-approaches}
@@
 The F\# type system also includes support for checking that units of measure are used correctly \cite{syme2012expert, kennedy1994dimension}, but this more specialized static invariant is left entirely unchecked at language boundaries.
-Exposing functions that operate over datatypes, tuples and values having units of measure is not recommended when a component ``might be used'' from another language, according to drafted F\# component design guidelines \cite{fsharpguidelines}, because it is awkward to construct and consume these from other languages without the convenient primitive operations (e.g.
+Exposing functions that operate over datatypes, tuples and values having units of measure is not recommended when a component ``might be used'' from another language \cite{syme2012expert} because it is awkward to construct and consume these from other languages without the convenient primitive operations (e.g.
 pattern matching) and syntax that F\# includes.
-SML.NET prohibits exposing such types at component boundaries altogether (and cannot naturally consume F\# data structures, despite having a rather similar syntax and semantics in many ways).
+SML.NET prohibits exposing such types at component boundaries altogether.
+It cannot it naturally consume F\# data structures, despite having a rather similar syntax and semantics in many ways (both languages directly descend from ML).
 %In Scala, traits that have default method implementations are difficult to implement from Java or other JVM languages and the workaround can break if the trait is modified \cite{scalatraitinterop}. 

### 2014-03-17 thesis-proposal/timeline.tex
Updated ICSE14 and proposal

--- 2014-01-16 thesis-proposal/timeline.tex
@@
 % !TEX root = omar-thesis-proposal.tex
+\section{Conclusion}\label{conclusion}
+In arguing for language extensibility, we are accepting Carnap's logical pluralism\todo{citations}. But we take issue with his \emph{principle of tolerance}, which he explained as follows:
+\begin{quote}
+In logic, there are no morals.
+Everyone is at liberty to build his own logic, i.e.
+his own form of language, as he wishes.
+All that is required of him is that, if he wishes to discuss it, he must state his methods clearly, and give syntactic rules instead of philosophical arguments.
+\end{quote}
+%Aside from the archaic use of gendered pronouns, 
+Carnap formulated his principle of tolerance because he wished to freely explore the consequences of reasoning with different connectives and  rules and emphasize that conducting philosophical reasoning by methodical logical analysis did not require one to accept any particular axioms.
+But he was also a prominent member of the Vienna circle, which sought a ``unified science'' the purpose of which was "to link and harmonise the achievements of individual investigators in their various fields of science" by the construction of a "constitutive system" \cite{ViennaCircle}.
+These two programs were in conflict.
+If the reader will permit a brief digression into social science, the adoption of a tolerant viewpoint is difficult to sustain in a society that does not require \emph{reciprocity}.
+Yossie Nehushtan goes as far as to argue that a liberal  society ``should not tolerate anything that denies the justifications of tolerance'' \cite{Nehushtan}.
+According to Raz, the primary justification of tolerance is to maximize
+individual autonomy in a multi-party context \cite{Raz}.
+And indeed, this was precisely Carnap's program.
+But were the Vienna circle's ``constitutive system'' to become na\"ively tolerant of new inference rules (what Carnap called L-rules), it would find that this would then deny its investigators autonomy of reasoning.
+Thus, we propose a more cooperative \emph{principle of reciprocal tolerance}:
+\begin{quote}
+In metalogic, there \emph{are} morals.
+Everyone is at liberty to build their own logical fragment, i.e.
+their own language extensions, as long as these can be clearly shown, by formal means instead of philosophical argument, to conserve autonomously derived reasoning principles when used in \emph{any} combination.
+\end{quote}
+
+By following this principle, we believe that a research program that supports Carnap's logical pluralism within a ``constitutive system'' of the sort envisioned by Vienna circle is both possible and practical.
+The work in this thesis gives first steps towards this goal.
+
+% http://www.trinitinture.com/documents/nehushtan.pdf
+% https://en.wikipedia.org/wiki/Vienna_Circle
+% http://consequently.org/papers/carnap.pdf
+% https://www.academia.edu/4041290/How_tolerant_can_you_be_Carnap_on_the_normativity_of_logic
+
 \section{Timeline}
+To summarize, we plan on completing the work in this thesis per the following schedule:
 \begin{itemize}
-\item @$\lambda$ - ICFP - Mar. 1
-\item Wyvern - ECOOP or OOPSLA - late March
-\item Ace - PLDI or OOPSLA - late March
-\item Then write everything up.
+\item The work on type-specific languages has been accepted to ECOOP 2014. The deadline for final revisions is May 12, so we will complete the remaining work by then.
+\item The work on Ace will be submitted to OOPSLA 2014 on Mar. 25.
+\item The work on @$\lambda$ and active embeddings is substantially complete.
+We plan to complete some technical details and exposition after Mar.
+25 and submit it to POPL in July (though we hope to be done with the thesis work before then).
+\item The work on active code completion has been completed and published at ICSE 2012 \cite{Omar:2012:ACC:2337223.2337324}.
+The remaining work is expository and will be completed in the course of writing the dissertation.
 \end{itemize}

### 2014-03-19 thesis-proposal/parsing.tex
proposal updates

@@
 \ref{f-example} with an example showing several different TSLs being used to define a fragment of a web application showing search results from a database.
-Note that for clarity of presentation, we color each character  according to the TSL it is governed by.% Black represents the base language.
+Note that for clarity of presentation, we color each character  according to the TSL it is governed by.
+Black represents the base language and comments are in italics.
 

### 2014-03-19 thesis-proposal/problem.tex
proposal updates

@@
 	\item out-of-bounds backreferences to a captured group are not used \cite{spishak2012type}
-	\item string processing operations do not lead to a string that is malformed, when well-formedness can be captured as membership in a regular language \cite{fulton-thesis}
+	\item string processing operations do not lead to a string that is malformed, when well-formedness can be captured as membership in a regular language \cite{fulton-thesis,HosoyaVouillonPierce2000ICFP}
 	\end{enumerate}

### 2014-03-23 ace-oopsla14/ace-oopsla14.tex
ace-oopsla14 commit

--- 2013-11-21 ace-pldi14/ace-pldi14.tex
@@
 %\documentclass[12pt]{article}
-\documentclass[9pt,preprint]{sigplanconf}
+\documentclass[10pt,preprint]{sigplanconf}
+\newcommand{\lamAce}{\lambda_{\text{Ace}}}
 % The following \documentclass options may be useful:
@@
 \definecolor{mauve}{rgb}{0.58,0,0.82}
-\definecolor{light-gray}{gray}{0.75}
+\definecolor{light-gray}{gray}{0.5}
 \usepackage{listings}
@@
   tabsize=4,
-  commentstyle=\itshape,
+  commentstyle=\itshape\color{light-gray},
   basicstyle=\ttfamily\scriptsize,
-  morekeywords={lambda, self, assert, as},
+  morekeywords={lambda, self, assert, as,type},
   numbers=left,
@@
 
-\title{\Ace: Growing a Statically-Typed Language Inside a Python}
+\title{\Ace: Growing A Statically-Typed Language Inside a Python}
 
 \authorinfo{~}{~}{~}
-%\authorinfo{Cyrus Omar\and Jonathan Aldrich}
+%\authorinfo{Cyrus Omar\and Nathan Fulton\and Jonathan Aldrich}
  %          {School of Computer Science\\
   %          Carnegie Mellon University}
-   %        {\{comar, jonathan.aldrich\}@cs.cmu.edu}
+   %        {\{comar, nfulton, aldrich\}@cs.cmu.edu}
 
@@
 %abstractions, all within libraries.
-
-Evidence suggests that programmers are reluctant to adopt new programming languages to gain access to new abstractions, even when they agree that these abstractions would be valuable to them.
-This suggests a need for languages that are  \emph{compatible} with existing languages, tools and infrastructure and \emph{internally extensible}, so that adopting a new  primitive abstraction requires only importing a library in the usual way.
-
-In this paper, we introduce Ace, a language compatible with tools and infrastructure developed for Python, one of the most  widely-adopted dynamically-typed languages today.
-While Python, like other similar languages, was designed for simple scripting tasks, Ace is designed for more complex situations where static typechecking and programmatic control over compilation may be beneficial.
-Unlike most statically-typed languages, however, Ace's type system and semantics can be extended from within by novel mechanisms that avoid key interference issues faced by previous mechanisms.
-We show that these can be used to implement a range of statically-typed functional, object-oriented, parallel, low-level and domain-specific abstractions, as well as safe interoperability layers with existing languages, as orthogonal libraries.
+Programmers are justifiably reluctant to adopt new language dialects to access stronger type systems.
+This suggests a need for a language that is \emph{compatible} with existing libraries, tools and infrastructure and that has an \emph{internally extensible type system}, so that adopting and combining type systems requires only importing libraries in the usual way, without the need to resolve link-time ambiguities.
+
+We introduce Ace, an extensible statically typed language embedded within and compatible with Python, a widely-adopted dynamically typed language.
+Python serves as Ace's type-level language.
+Rather than building in a particular type system, the Ace compiler inverts control over type assignment and translation to type-level functions according to a type-directed protocol that does not lead to ambiguities when type systems are combined.
+We show that this protocol is flexible enough to admit library-based implementations of types  drawn from functional, object-oriented, parallel, low-level and domain-specific languages.
+We also show how this permits the construction of safe, natural and extensible foreign function interfaces using a novel form of ``staged'' type propagation from Python into Ace.
+We also give a simplified calculus that describes how a type-level language can be used to enable type system extensions and demonstrates how type safety can be maintained by lifting a technique from the typed compilation literature into the language.%To show that the method scales, we give a full implementation of the OpenCL programming language as a library.
+%We briefly examine the type theoretic foundations of this approach and give modular criteria that guarantee that an extension is safe and modular.
 %Ace can be used from the shell or interactively from within Python\todo{take this sentence out?}.
@@
 \section{Introduction}\label{intro}
-Asking programmers to import a new library is far more practical than asking them to  adopt a new programming language.
-Indeed, recent empirical studies underscore  the difficulties of driving new languages into adoption, finding that extrinsic factors like compatibility with existing codebases and libraries, team familiarity and tool support are at least as important as intrinsic factors \cite{Meyerovich:2013:EAP:2509136.2509515,chen05,nguyen2010survey}.
-As a result, many developers cannot use  abstractions they might prefer because these abstractions are only available in languages they cannot adopt \cite{Meyerovich:2012:SDR:2414721.2414724,Meyerovich:2013:EAP:2509136.2509515}.
-This issue was perhaps most succinctly expressed by a participant in a recent study by Basili et al.
+Asking programmers to import a new library is simpler than asking them to adopt a new programming language.
+Indeed, recent empirical studies underscore the difficulties of driving languages into adoption, finding that extrinsic factors like compatibility with large existing code bases,  library support, team familiarity and tool support are at least as important as intrinsic features of the language  \cite{Meyerovich:2013:EAP:2509136.2509515,chen05,nguyen2010survey}.
+
+Unfortunately, researchers and domain experts (\emph{providers}) designing potentially useful new abstractions can sometimes find it difficult to implement them as libraries, particularly when they require strengthening a language's type system.
+In these situations, abstraction providers often develop a new language or language dialect.
+Unfortunately, this \emph{language-oriented approach} \cite{language-oriented} does not scale to large applications: using  components written in languages with different type systems can be awkward and lead to safety problems and performance overhead at interface boundaries.
+%To avoid this, it is necessary to combine the different languages and their associated tools, but this often requires significant effort and expertise (we will discuss this further in Sec.
+\ref{related-work}).
+
+This is a problem even when a dialect introduces only a small number of new constructs.
+For example, a recent  study \cite{cave2010comparing} comparing a Java dialect, Habanero-Java (HJ), with a comparable library, \verb|java.util.concurrent|, found that the language-based abstractions in HJ were easier to use and provided useful static guarantees.
+Nevertheless, it concluded that the library-based abstractions remained more practical outside the classroom because HJ, as a distinct dialect of Java with its own type system, would be difficult to use in settings where some developers had not adopted it, but needed to interface with code that had adopted it.
+It would also be difficult to use it in combination with other abstractions also implemented as dialects of Java.
+Moreover, its tool support is more limited.
+This suggests that today, programmers and development teams cannot use  the abstractions they might prefer because they are only available bundled with languages they cannot adopt \cite{Meyerovich:2012:SDR:2414721.2414724,Meyerovich:2013:EAP:2509136.2509515}.
+
+ %This issue was perhaps most succinctly expressed by a participant in a recent study by Basili et al.
 \cite{basili2008understanding} who stated ``I hate MPI, I hate C++.
 [But] if I had to choose again, I would probably choose the same.''
-
-Unfortunately, researchers and domain experts who design and develop potentially useful new abstractions can find it difficult to implement them in terms of the general-purpose abstraction mechanisms available in mainstream languages.
-This is particularly salient for abstractions that require support from the typechecker or compiler, such as those focused on correctness and performance, as well as those that  introduce more concise or natural notations for existing abstractions.
-For example, a recent controlled study comparing a new language, Habanero-Java (HJ), with a comparable library, \verb|java.util.concurrent|, found that the language-based approach was more concise, correct and easy-to-use, but concluded that the library-based approach was more practical outside the classroom because HJ introduced new constructs and keywords into Java, requiring the adoption of a new toolchain, which could lead to compatibility issues with plain Java code \cite{cave2010comparing}.
-
-Internally-extensible languages promise to reduce the need for new standalone languages by giving abstraction providers more direct control over a base language's syntax and semantics from within libraries.
-By using such a language, programmers  gain the ability to granularly import the primitive abstractions best suited to each part of their application or library.
-The research community thus gains the ability to more easily develop, deploy and evaluate new abstractions in the context of existing codebases, narrowing one of the gaps between research and practice \cite{basili2008understanding}.
-
+%Unfortunately, researchers and domain experts (collectively, \emph{providers}) who design potentially useful new abstractions often find it difficult to implement them in terms of the general-purpose abstraction mechanisms available in existing languages.
+%Using a foreign construct can be  unnatural and, if the languages have  different type systems, can also cause safety problems.
+
+Internally extensible languages promise to reduce the need for new dialects by giving abstraction providers more direct control over a language's syntax and static semantics from within libraries.
+%In   an internally extensible language, programmers  can \emph{granularly} import the primitive constructs best suited to each component of their application or library.
+Providers can thus more easily develop, deploy and evaluate new abstractions in the context of existing codebases, narrowing one of the gaps between research and practice \cite{basili2008understanding}.
 %Indeed, abstractions could compete on their individual merit, rather than  by the serious adoption barriers mentioned above.
-
-Unfortunately, internally-extensible languages available today have several problems.
-First, an extension mechanism itself may require modifying a base language with constructs for defining, importing and using extensions.
-The extension mechanism is not itself a library, so it faces a ``bootstrapping'' problem: it must overcome many of the same extrinsic issues as other new languages like HJ.
-The extension mechanisms available today also have some intrinsic problems related to safety and expressiveness that require technical solutions before it would be appropriate to widely rely on them.
-
-This paper describes the design and implementation of Ace, an internally-extensible language designed considering both extrinsic and intrinsic criteria.
-To solve the bootstrapping problem, Ace is implemented entirely as a library within the popular Python programming language.
-Ace and Python share a common syntax and package system, allowing Ace to leverage its well-established tools and infrastructure directly.
-Python serves as the compile-time metalanguage for Ace, but Ace functions themselves do not operate according to Python's fixed dynamically-typed semantics  (cf.
-\cite{Politz:2013:PFM:2509136.2509536,python}).
-Instead, Ace has a statically-typed semantics that can be extended by users from within libraries.
-
-More specifically, each Ace function can be annotated with a base semantics that determines the meaning of simple expressions like literals and certain statements.
-The semantics of the remaining expressions and statements are governed by logic associated with the type of a designated subexpression.
-We call the user-defined base semantics \emph{active bases} and the types in Ace \emph{active types}, borrowing terminology from \emph{active libraries} (\cite{activelibraries}, see Sec.
+Unfortunately, the mechanisms available today have several problems.
+First, they are themselves often available only within a dialect of an existing language and thus face a ``chicken-and-egg'' problem: a language like SugarJ \cite{sugarj} must overcome the same extrinsic issues as a language like HJ.
+Second, giving providers too much control over the language can lead to intrinsic safety issues, ambiguities and conflicts between extensions, as we will discuss.
+Too little control, on the other hand, leaves it difficult to implement interesting abstractions.%It is difficult to guarantee that extensions are sound in isolation and more so to guarantee that they can be {safely} combined.
+%This requires further research before it would be appropriate to widely rely on them (which we will discuss as we go on).
+
+This paper describes the design and implementation of Ace, an internally extensible, statically typed programming language designed considering both these extrinsic and intrinsic factors.
+To address the ``chicken-and-egg'' problem, Ace is implemented within Python, entirely as a library  \cite{Politz:2013:PFM:2509136.2509536,python}.
+The top-level of an Ace file can be thought of as a \emph{compilation script} written in Python and, as we will discuss, Python serves as Ace's type-level language.
+Ace and Python share a common syntax and package system, so users of Ace can leverage established tools and infrastructure directly.
+%The semantics of  define run-time behavior.
+
+Functions marked by a decorator as under the control of Ace are \emph{statically typechecked}.
+This makes it possible to rule out many potential issues ahead of execution and thus avoid the need for costly dynamic checks.
+% Instead, Ace has a statically-typed semantics that can be extended by users from within libraries.
+However, there is \emph{no particular type system built into Ace}.
+Instead, a type is any  {compile-time object} implementing the \verb|ace.Type| interface.
+We call these \emph{active types}.
+When statically assigning a type to a compound form, e.g.
+\verb|e.x|, the Ace {compiler} delegates to the type of a subexpression, e.g.
+to the type of \verb|e|, to determine how to assign a type to the expression as a whole.
+As we will show, this {type-directed protocol}, which we call \emph{active type assignment}, permits the expression of a rich variety of type systems as libraries.
+Unlike syntax-directed protocols, there cannot be ambiguities when separately defined type sytems are combined (extensions have exclusive control over non-overlapping sets of expressions by construction).
+Base cases, e.g.
+variables and literals, are handled on a {per-function basis} by a \emph{base semantics}, also a compile-time object implementing an interface, \verb|ace.Base|.
+
+% As a result, clients are able to import any combination of extensions with the confidence that link-time ambiguities cannot occur .
+
+The dynamic behavior of an Ace function is determined by translation to a \emph{target language}.
+We begin by  targeting Python, then discuss targeting OpenCL and CUDA, lower-level languages used to program many-core processors (e.g.
+GPUs).
+Active types and bases control translation by the same type-directed protocol that governs type assignment.
+We call this phase \emph{active translation}.
+A \emph{target} mediates this process and is also a user-defined compile-time object implementing an interface, \verb|ace.Target|.
+%One base or type can support multiple targets. 
+When targeting a typed language, care must be taken to ensure that well-typed Ace expressions have well-typed translations, so we integrate a technique for compositionally reasoning about compiler correctness (developed for the TIL compiler for Standard ML \cite{TIL}) into the language.
+
+Ace can be used from the shell, producing source files that can be further compiled or executed by external means.
+Ace functions targeting a language with Python bindings (e.g.
+Python itself, as well as the others just described) can also be compiled and invoked interactively from Python scripts.
+Ace supports a form of \emph{ad hoc polymorphism} (based on singleton types) whereby types propagate into generic functions.
+This occurs both between Ace functions and when calling Ace functions from Python.
+In the latter case, the ``dynamic type'' of a Python value determines a static type ``just-in-time'' when a call occurs.
+This represents a form of implicit \emph{staging} and significantly streamlines interactive workflows (common in, for example, scientific computing) that alternate between Python for orchestration and productivity-oriented tasks (e.g.
+exploratory plotting) and an external language via a foreign-function interface (FFI) for performance- and correctness-critical portions.%We will discuss in detail as our primary realistic case study.
+
+The remainder of the paper is organized as follows: in Sec.
+\ref{usage}, we describe the basic structure and usage of Ace with an example that simultaneously uses dynamically typed terms (which can be thought of as having the static type \verb|dyn|) alongside statically typed immutable records and a simple prototypic object system, all implemented orthogonally as active types.
+In Sec.
+\ref{att}, we detail how active type assignment and translation delegate control, at compile-time, to active types and bases.
+In Sec.
+\ref{theory}, we reduce this mechanism to  a core lambda calculus, $\lamAce$,  clarify the relationship to   type-level computation and show how type safety is maintained.
+In Sec.
+\ref{targets}, we discuss generic functions, FFIs and staging via a complete implementation of the OpenCL type system (and extensions to it) as a library.
+In Sec.
+\ref{related}, we compare Ace to related work.
+We conclude in Sec.
+\ref{discussion} by summarizing our contributions, highlighting the key features needed by a host language to support these mechanisms, and discussing limitations and potential future work.
+The appendix contains details omitted from the main body and additional examples.
+
+\section{Language Design and Usage}\label{usage}
+
+%
+Listing \ref{example} shows an example of an Ace compilation script.
+As promised, compilation scripts are written directly in Python.
+Ace requires no modifications to the language or features specific to the primary implementation, CPython (so Ace supports alternative implementations like Jython and PyPy).
+This choice pays immediate dividends on the first five lines: Ace's package system is Python's package system, so Python's build tools (e.g.
+\verb|pip|) and package repostories (e.g.
+\verb|PyPI|) are directly available for distributing libraries that use Ace.
+Note that all of the type systems that we discuss are implemented entirely as  libraries.
+
+%The top-level statements in an Ace file, like the \verb|print| statement on line 10, are executed to control the compile-time behavior, rather than the run-time behavior, of the program.
+%That is, Python serves as the \emph{compile-time metalanguage} (and, as we will see shortly, the \emph{type-level language}) of Ace.
+%For readers familiar with C/C++, Python can be thought of as serving a role similar to (but more general than) its preprocessor and template system (as we will see).
+
+\subsection{Types}
+\begin{codelisting}
+\begin{lstlisting}
+from examples.py import py, string
+from examples.fp import record
+from examples.oo import proto
+from examples.num import decimal
+from examples.regex import string_in
+
+A = record['amount' : decimal[2]]
+C = record[
+  'name' : string, 
+  'account_num' : string_in[r'\d{10}'],
+  'routing_num' : string_in[r'\d{2}-\d{4}/\d{4}']]
+Transfer = proto[A, C]
+
+@py
+def log_transfer(t):
+  """Logs a transfer to the console."""
+  {t : Transfer}
+  print "Transferring %s to %s." % (
+    [string](t.amount), t.name)
+
+@py
+def __toplevel__():
+  print "Hello, run-time world!"
+  common = {name: "Annie Ace", 
+            account_num: "0000000001", 
+            routing_num: "00-0000/0001"} (C)
+  t1 = ({amount: 5.50}, common) (Transfer)
+  t2 = ({amount: 15.00}, common) (Transfer)
+  log_transfer(t1)
+  log_transfer(t2)
+
+print "Hello, compile-time world!"
+\end{lstlisting}
+\caption{[\texttt{listing\ref{example}.py}] An Ace compilation script.}
+\label{example}
+\end{codelisting}
+
+Types are constructed programmatically during execution of the compilation script.
+This stands in contrast to many  contemporary statically-typed languages, where types (e.g.
+datatypes, classes, structs) can only be declared.
+Put another way, Ace supports \emph{type-level computation} \cite{tlc} and Python is its type-level language (discussed further in Sec.
+\ref{theory}).
+%The compilation script can assign shorter names to types for convenience (removing the need for  facilities like \verb|typedef| in C or \verb|type| in Haskell). 
+In our example, we see several types being constructed:
+\begin{enumerate}
+\item On line 7, we construct a functional record type with a single (immutable) field named \verb|amount| with type \verb|decimal[2]|, which classifies decimal numbers with two decimal places.
+Here, \verb|record| and \verb|decimal| are \emph{indexed type constructors}.
+We will see how they are implemented in Sec.
+\ref{att}, but note that syntactically, \verb|record| borrows Python's syntax for array slices (e.g.
+\verb|a[min:max]|).
+Note that the field name could equivalently have been written \lstinline{'a' + 'mount'}, again emphasizing that types and indices are constructed programatically by a Python script.
+\item On lines 8-11, we construct another record type.
+The field \verb|name| has type \verb|string|, defined in \verb|examples.py|, while the fields \verb|account_num| and \verb|routing_num| have more interesting types, classifying strings guaranteed statically to be in the language of a regular expression  (written, to avoid needing to escape backslashes, using Python's \emph{raw string literals}).
+Our appendix specifies the typechecking rules, based on \cite{nfulton-thesis}.
+We include it to emphasize that we aim to support specialized type systems,  not just general purpose constructs like records and numbers.
+\item Prototypic inheritance is a feature of some dynamically-typed languages (most notably, Self \cite{self} and Javascript).
+On line 12, we consider a simple statically-typed variant.
+The type \verb|Transfer| classifies terms consisting of a \todo{what is the right term for this?}\emph{fore} of type \verb|A| and a \emph{prototype} of type \verb|C|.
+If a field cannot be found in the fore, the type system will delegate (here, statically) to the prototype.
+This makes it easy to share the values of the fields of a common prototype amongst many fores.
+In our example, we will perform multiple transfers differing only in amount.
+\end{enumerate}
+\subsection{Functions}
+Functions implementing run-time behavior are distinguished by a decorator specifying their  \emph{base semantics} (here \verb|py| from \verb|examples.py|; we will discuss this further below).
+These functions are, however, still written using Python's syntax, a choice that is again valuable for extrinsic reasons: users of Ace can use a variety of  tools designed to work with Python source code  without  modification, including code highlighters (like the one used in generating this paper), editor plugins, style checkers and documentation generators.
+We will see how, with a bit of cleverness, Python's syntax can support a variety of static semantics.
+\begin{codelisting}
+\begin{lstlisting}[style=Bash]
+$ `acec listing1.py`
+Hello, compile-time world!
+[acec] _listing1.py successfully generated.
+$ `python _listing1.py`
+Hello, run-time world!
+Transferring 5.50 to Annie Ace.
+Transferring 15.00 to Annie Ace.
+\end{lstlisting}
+\caption{Compiling \texttt{listing\ref{example}.py} using \texttt{acec}.
+Both steps can be performed at once by writing \lstinline[style=Bash]{`ace listing1.py`} (line 3 will not be printed with this command).}
+\label{external-compilation}
+\end{codelisting}
+\begin{codelisting}[t]
+\begin{lstlisting}
+import examples.num.runtime as __ace_0
+
+def print_transfer(t):
+  print "Transferring %s to %s." % (
+    __ace_0.decimal_to_str(t[0],2), t[1][0])
+
+print "Hello, run-time world!"
+common = ("Annie Ace", "0000000001", "00-0000/0000")
+t1 = ((5,50), common)
+t2 = ((15,0), common)
+print_check(t1)
+print_check(t2)
+\end{lstlisting}
+\caption{[\texttt{\_listing\ref{example}.py}] The file generated in Listing \ref{external-compilation}.}
+\label{example-out}
+\end{codelisting}
+
+On lines 14-18, we see the function \verb|log_transfer|.
+As with a Python function, we write a documentation string like the one shown specifying, informally, the behavior of the function.
+Before moving into the body, however, we  also write a \emph{type signature} (line 17) stating simply that the type of \verb|t| is \verb|Transfer|, the prototypic object type described above.
+As a result, we can assume that \verb|t.amount| and \verb|t.name| have types \verb|decimal[2]| and \verb|string|, respectively.
+We will return to the details of \verb|print| and the form \verb|[string](t.amount)|, which performs an explicit conversion, shortly.\todo{when?}
+
+Line 17 borrows Python's syntax for dictionary literals to approximate  conventional notation for type annotations.\footnote{If the right-hand side of every annotation is not a type, it will be an error.
+In the rare situation where there are no type annotations and the first statement is a dictionary literal, users can use Python's \texttt{pass} keyword to skip the line.} % (and following conventions for documentation strings, which also describe ``types'' at the end).
+% The ``keyword'' \verb|type| was chosen because it is the name of a built-in function in Python (and thus would not be used in the way shown for any other purpose).
+ {In version 3.0 of Python, syntax for annotating arguments with arbitrary metadata using \verb|:| was introduced,  intended for use by future tools (further suggesting that users of dynamically typed languages see potential in a static typing discipline)  \cite{PEPwhatever}.
+Ace supports both notations when the compilation script is run using Python 3.\textit{x}, but we use the more universally available notation here to satisfy our extrinsic goals: the Python 2.\text{x} series  remains the most widely used by a large margin as of this writing (we will consider cross-compilation in Sec.
+\ref{targets}, however).
+
+\subsection{Introductory Forms}
+On lines 21-30, we define the function \verb|__toplevel__| (the base will consider this a special name for the purposes of compilation, discussed in the Sec.
+\ref{compilation}).
+After printing a run-time greeting, we construct a value of type \verb|C| on lines 24-26.
+Recall that \verb|C| is a record type, so we provide the names of the fields (here, without quotes because we are no longer in the type-level langauge) and values of the appropriate type using syntax for a dictionary literal.
+Because there may be multiple types that use this form, we say which type the literal should be \emph{analyzed} against with the \emph{literal ascription}  \verb|(C)|.
+This is another repurposing of Python's syntax: here function application.
+When the ``function'' has one of Python's literal forms and the (single) argument is a type, we treat it instead as an \emph{ascribed introductory form}.
+The literal is treated as an introductory form for that type (we will see precisely how this occurs shortly).
+We see the same form used with tuple literals on lines 27-28 to introduce terms of type \verb|Transfer|.
+
+\subsection{Bidirectional Typechecking}
+The string, number and dictionary forms inside the outermost forms on lines 24-28 are also introductory forms, but we do not include an ascription because the definition of \verb|C| determines which types they should be analyzed against.
+As we will discuss shortly, Ace has a \emph{bidirectional type system} distinguishing locations where a type must be \emph{synthesized} from an expression (e.g.
+at the top-level) and locations where an expression is \emph{analyzed} against a type \cite{LovasPfenning08}.
+The base determines the type a literal should be analyzed against in a synthetic context.
+For example, if we did not specify the literal ascription \verb|(C)| on line 26, the base would analyze it against the Python dictionary type, which would cause a type error because the variables used as keys were not defined in the surrounding context.
+
+\subsection{Compilation}\label{compilation}
+
+%Because \verb|t| will be statically guaranteed to have type \verb|Transfer|,  
+\subsection{B}
+\section{Active Type Synthesis and Translation}\label{att}
+The core of Ace consists of about 1500 lines of Python code implementing its primary concepts: generic functions, concrete functions, active types, active bases and active targets.
+The latter three comprise Ace's extension mechanism.
+Extensions provide semantics to, and govern the compilation of, Ace functions, rather than logic in Ace's core.
+%Indeed, the name ``Ace'' might itself be an acronym: it is an ``active compilation environment''.
+
+\subsection{Active Types}\label{atypes}
+Active types are the primary means for extending Ace with new abstractions.
+An active type, as mentioned previously, is an instance of a class implementing the \verb|ace.ActiveType| interface.
+Listing \ref{cplx} shows an example of such a class: the \verb|clx.Cplx| class used in Listing \ref{compscript}, which implements the logic of complex numbers.
+The constructor takes as a parameter any numeric type in \verb|clx| (line \ref{cplx}.2).
+
+\subsubsection{Dispatch Protocol}
+\begin{codelisting}
+\lstinputlisting{listing9.py}
+\caption{\texttt{[}in \texttt{examples/clx.py]} The active type family \texttt{Ptr} implements the semantics of OpenCL pointer types.}
+\label{cplx}
+\end{codelisting}
+In a compiler for a monolithic language, there would be a \emph{syntax-directed} protocol governing typechecking and translation.
+In a compiler written in a functional language, for example, one would declare datatypes that captured all forms of types and expressions, and the typechecker would perform exhaustive case analysis over the expression forms.
+That is, all the semantics are implemented in one place.
+The visitor pattern typically used in object-oriented languages implements essentially the same protocol.
+This does not work for an extensible language because new cases and logic need to be added \emph{modularly}, in a safely composable manner.
+
+Instead of taking a syntax-directed approach, the Ace compiler's typechecking and translation phases take a \emph{type-directed approach}.
+When encountering a compound term (e.g.
+\verb|e[e1]|), the compiler defers control over typechecking and translation to the active type of a designated subexpression (e.g.
+\verb|e|) determined by Ace's fixed \emph{dispatch protocol}.
+Below are examples of the choices made in Ace.
+Due to space constraints, we do not show the full protocol.
+\begin{itemize}
+\item Responsibility over {\bf attribute access} (\texttt{e.attr}), {\bf subscripting} (\texttt{e[e1]}) and \textbf{calls} (\verb|e(e1, ..., en)|) and {\bf unary operations} (e.g.
+\verb|-e|) is handed to the type recursively assigned to \texttt{e}.
+\item Responsibility over {\bf binary operations} (e.g.
+\verb|e1 + e2|) is first handed to the type assigned to the left operand.
+If it indicates a type error, the type assigned to the right operand is handed responsibility, via a different method call.
+{Note that this operates like the corresponding rule in Python's \emph{dynamic} operator overloading mechanism; see Sec.
+\ref{related} for a discussion.}
+\item Responsibility over \textbf{constructor calls} (\verb|[t](e1, ..., en)|), where \verb|t| is a \emph{compile-time Python expression} evaluating to an active type, is handed to that type.
+If \verb|t| evaluates to a \emph{family} of types, like \verb|clx.Cplx|, the active type is first generated via a class method, as discussed below.
+%\item Responsibility over {\bf simple assignment statements}, is handed to the type of the variable on the left (which, as we will see below, is determined by the active base).
+If this type does not provide special assignment semantics, the base must handle it.%Destructuring assignment is also supported by a somewhat more complex protocol.\todo{clarify}
+\end{itemize}
+
+\subsubsection{Typechecking}
+When typechecking a compound expression or statement, the Ace compiler temporarily hands control to the object selected by the dispatch protocol by calling the method \verb|type_|$X$, where $X$ is  the name of the syntactic form, taken from the Python grammar \cite{python} (appended with a suffix in some cases).
+
+For example, if \verb|c| is a complex number, then \verb|c.ni| and \verb|c.i| are its non-imaginary and imaginary components, respectively.
+These expressions are of the form \verb|Attribute|, so the typechecker calls \verb|type_Attribute| (line \ref{cplx}.7).
+This method receives the compilation context, \verb|context|, and the abstract syntax tree of the expression, \verb|node| and must return a type assignment for the node, or raise an \verb|ace.TypeError| if there is an error.
+In this case, a type assignment is possible if the attribute name is either \verb|"ni"| or \verb|"i"|, and an error is raised otherwise (lines \ref{cplx}.8-\ref{cplx}.10).
+We note that error messages are an important and sometimes overlooked facet of {ease-of-use} \cite{marceau2011measuring}.
+A common frustration with using general-purpose abstraction mechanisms to encode an abstraction is that they can produce  verbose and cryptic error messages that reflect the implementation details instead of the semantics.
+Ace supports custom error messages.%Active types permit direct encoding of semantics and customization of error messages.
+
+Complex numbers also support binary arithmetic operations partnered with both other complex numbers and with non-complex numbers, treating them as if their imaginary component is zero.
+The typechecking rules for this logic is implemented on lines \ref{cplx}.17-\ref{cplx}.29.
+Arithmetic operations are usually symmetric, so the dispatch protocol checks the types of both subexpressions for support.
+To ensure that the semantics remain deterministic in the case that both types support the binary operation, Ace asks the left first (via \verb|type_BinOp_left|), asking the right (via \verb|type_BinOp_right|) only if the left indicates an error.
+In either position, our implementation begins by recursively assigning a type to the other operand in the current context via the \verb|context.type| method (line \ref{cplx}.24).
+If supported, it applies the C99 rules for arithmetic operations to determine the resulting type (via \verb|c99_binop_t|, not shown).
+
+Finally, a complex number can be constructed inside an Ace function using Ace's special constructor form: \verb|[clx.Cplx](3,4)| represents $3+4i$, for example.
+The term within the braces is evaluated at \emph{compile-time}.
+Because \verb|clx.Cplx| evaluates not to an active type, but to a class, this form is assigned a type by handing control to the class object via the \emph{class method} \verb|type_New|.
+It operates as expected, extracting the types of the two arguments to construct an appropriate complex number type (lines \ref{cplx}.50-\ref{cplx}.57), raising a type error if the arguments cannot be promoted to a common type according to the rules of C99 or if two arguments were not provided.% (an exercise for the reader: modify this method to also allow a single argument for when the imaginary part is 0).
+
+%Note that in each example, subexpressions are not \emph{automatically} assigned types.
+Types must be requested from the context.
+This is useful because it allows types to reinterpret subexpressions of terms they control, to supporting more complex constructs (e.g.
+pattern matching on functional datatypes, see Sec.
+\ref{datatypes}).
+
+\subsubsection{Translation}
+Once typechecking a method is complete, the compiler enters the translation phase, where terms in the target language are generated from Ace terms.
+Terms in the target language are generated by calling methods of the \emph{active target} governing the function being compiled.
+The translation phase operates similarly to typechecking, using the dispatch protocol to invoke methods named \verb|trans_|$X$.
+These methods have access to the context and node just as during typechecking, as well as the active target (named \verb|target| here).
+
+As seen in Listing \ref{mapout}, we are implementing complex numbers internally using OpenCL vector types, like \verb|int2|.
+Let us look first at \verb|trans_New| on lines \ref{cplx}.54-\ref{cplx}.60, where new complex numbers are translated to vector literals by invoking \verb|target.VecLit|.
+This will ultimately  generate the necessary OpenCL code, as a string, to complete compilation (these strings are not directly manipulated by extensions, however, to avoid problems with, e.g.
+precedence).
+For it to be possible to reason compositionally about the correctness of compilation, all complex numbers must translate to terms in the target language that have a consistent target type.
+The \verb|trans_type| method of the \verb|ace.ActiveType| associates a type in the target language, here a vector type like \verb|int2|, with the active type.
+Ace supports a mode where this \emph{representational consistency} is dynamically checked during compilation (requiring that the active target know how to assign types to terms in the target language, which can be done  for our OpenCL target as of this writing).
+
+The translation methods for attributes (line \ref{cplx}.12) and binary operations  (line \ref{cplx}.31) proceed in a straightforward manner.
+The context provides a method, \verb|trans|, for recursively determining the translation of subexpressions as needed.
+Of note is that the translation methods can assume that typechecking succeeded.
+For example, the implementation of \verb|trans_Attribute| assumes that if \verb|node.attr| is not \verb|'ni'| then it must have been \verb|'i'| on line \ref{cplx}.14, consistent with the implementation of \verb|type_Attribute| above it.
+Typechecking and translation are separated into two methods to emphasize that typechecking is not target-dependent, and to allow for more advanced uses, like type refinements and hypothetical typing judgements, that we do not describe here.
+
+\subsection{Active Bases}\label{abases}
+Each generic function is associated with an active base, which is an object implementing the \verb|ace.ActiveBase| interface.
+The active base specifies the \emph{base semantics} of that function.
+It controls the semantics of statements and expressions that do not have a clear ``primary'' subexpression for the dispatch protocol to defer to.
+A base is handed control over typechecking of statements and expressions in the same way as active types: via \verb|type_|$X$ and \verb|trans_|$X$ methods.
+Each function can have a different base.
+
+Literals are the most prominent form given their semantics by an active base.
+Our example active base, \verb|clx.base|, assigns integer literals the type \verb|clx.int32| while floating point literals have type \verb|clx.double|, consistent with the semantics of OpenCL and C99.
+The \verb|clx.base| object is an instance of \verb|clx.Base| that is pre-constructed for convenience.
+Alternative bases can be generated via different constructor arguments.
+For example, \verb|clx.Base(flit_t=clx.float)| is a base semantics where floating point literals have type \verb|clx.float|.
+This is useful because some OpenCL devices do not support double-precision numbers, or impose a significant performance penalty for their use.
+Indeed, to even use the \verb|double| type in OpenCL, a flag must be toggled by a \verb|#pragma| (The OpenCL target we have implemented inserts this automatically when needed, along with some other flags, and annotations like \verb|kernel|).
+Similarly, for some applications, avoiding accidental overflows is more important than performance.
+Infinite-precision integers can be implemented as an active type (not shown) and the base can be asked to use it for numeric literals.
+In all cases, this occurs by inserting a constructor call form, as described above.% the constructor call form, described above, around integers to support this.
+
+The base also initializes the context and governs the semantics of variables and assignment.
+This can also permit it to allow for the use of ``built-in'' operators that were not explicitly imported.
+The semantics of \verb|get_global_id| and \verb|printf| are provided by \verb|clx.base|.
+In fact, the base provides extended versions of them (the former permits multi-dimensional global IDs, the latter supports typechecking format strings).A base which does not provide them as built-ins (requiring that they be imported explicitly) can be generated by passing the \verb|primitives=False| flag.
+
+\subsection{Active Targets and Cross-Compilation}\label{atargets}
+An active target, which we describe being used above, is an instance of \verb|ace.ActiveTarget|.
+It has two primary roles in Ace: 1) providing an API for code generation to a target language, providing term and type constructors for use during translation (and, optionally, a type system implementation to support representational consistency checks); 2) implementing the interoperability layer to support calling of  generic or concrete function directly from Python, as described in Sec.
+\ref{implicit}.
+
+An active type or base can support multiple active targets by simply examining \verb|target| during translation (e.g.
+by performing capability checks using Python's \verb|hasattr| function).
+Alternatively, the active target can simulate the interface of a different target but generate code differently.
+For example, although our implementation of OpenCL's type system as a collection of active types and an active base relies on a target that supports OpenCL terms and types, our \verb|CUDA| and \verb|C99| targets provide partial support for the same term and type forms, in order to support cross-compilation.
+We will show the latter being used in Sec.
+\ref{examples}.
+
+\subsection{Compositional Reasoning}\label{safety}
+Every statement and expression in an Ace function is governed by exactly one active type, determined by the dispatch protocol, or by the single active base associated with the Ace function.
+The representational consistency check ensures that compilation is successful without requiring that types that abstract over other types (e.g.
+container types) know about their internal implementation details.
+Together, this permits compositional reasoning about the semantics of Ace functions even in the presence of many extensions.
+This stands on contrast to many prior approaches to extensibility, where extensions could insert themselves into the semantics in conflicting ways, making the order of imports matter (see Sec.
 \ref{related}).
-Both are objects that can be defined and manipulated at compile-time using Python.
-An important consequence of this mechanism is that it permits \emph{compositional} reasoning -- active bases and active types govern only specific non-overlapping portions of a program.
-As a result, clients are able to import any combination of extensions with the confidence that link-time ambiguities cannot occur (unlike many previous approaches, as we discuss in Sec.
-\ref{related}).
-
-The \emph{target} of compilation is also user-defined.
-We will show examples of Ace targeting Python as well as OpenCL, CUDA and C99, lower-level languages often used to program hardware.
-An active base or type can support multiple \emph{active targets}, which mediate translation of Ace code to code in a target language.
-Ace functions targeting a language with Python bindings can be called directly from Python scripts, with compilation occurring implicitly.
-For some data structures, types can propagate from Python into Ace.
-We show how this can be used to streamline the kinds of interactive workflows that Python is often used for.
-Ace can also be used non-interactively from the shell, producing source files that can be further compiled and executed by external means.
-
-The remainder of the paper is organized as follows: in Sec.
-\ref{usage}, we describe the basic structure and usage of Ace with an example library that internalizes and extends the OpenCL language.
-Then in Sec.
-\ref{att}, we show how this and other libraries are implemented by detailing the extension mechanisms within Ace.
-To explain and demonstrate the expressiveness of these mechanisms (in particular, active types) further, we continue in Sec.
-\ref{examples} by showing a diverse collection of abstractions drawn from different language paradigms that can be implemented as orthogonal libraries in Ace.
-We include functional datatypes, objects, and several other abstractions.
-In Sec.
-\ref{related}, we compare Ace to related work on language extensibility and metaprogramming.
-We conclude in Sec.
-\ref{discussion} by summarizing our contributions, discussing the essential features needed by a host language to support these mechanisms, and describing their limitations and potential future work.
-
-\section{Language Design and Usage}\label{usage}
-Listing \ref{map} shows an example of an Ace file.
-As promised, the top level of an Ace file is written directly in Python, requiring no modifications to the language (versions 2.6+ or 3.3+) nor features specific to CPython (so Ace supports alternative implementations like Jython, IronPython and PyPy).
-This choice pays immediate dividends on line 1: Ace's package system is Python's package system, so Python's build tools (e.g.
-\verb|pip|) and package repostories (e.g.
-\verb|PyPI|) are directly available for distributing Ace libraries.
-
-The top-level statements in an Ace file, like the \verb|print| statement on line 10, are executed to control the compile-time behavior, rather than the run-time behavior, of the program.
-That is, Python serves as the \emph{compile-time metalanguage} (and, as we will see shortly, the \emph{type-level language}) of Ace.
-%For readers familiar with C/C++, Python can be thought of as serving a role similar to (but more general than) its preprocessor and template system (as we will see).
-Functions containing run-time behavior, like \verb|map|, are governed by a semantics that can differ from Python's (in ways that we will describe below), but they share Python's syntax.
-As a consequence, users of Ace  immediately benefit also from an ecosystem of well-developed tools that work with Python syntax, including parsers, code highlighters, editor modes, style checkers and documentation generators.
-
+
+\section{Theoretical Foundations}\label{theory}
+\section{Generic Functions, Targets and Interactive Execution}\label{targets}
 \subsection{OpenCL as an Active Library}
@@
 \subsection{Generic Functions}\label{genfn}
-\begin{codelisting}
-\lstinputlisting[linebackgroundcolor={\btLstHL{5-8}}]{listing3.py}
-\caption{[\texttt{listing\ref{map}.py}] A generic data-parallel higher-order map function targeting OpenCL.}
-\label{map}
-\end{codelisting}
-\begin{codelisting}[t]
-\lstinputlisting[commentstyle=\color{mauve}]{listing7.py}
-\caption{[\texttt{listing\ref{metaprogramming}.py}] Metaprogramming with Ace, showing how to construct generic functions from abstract syntax trees.}
-\label{metaprogramming}
-\end{codelisting}
+%\begin{codelisting}[t]
+%\lstinputlisting[commentstyle=\color{mauve}]{listing7.py}
+%\caption{[\texttt{listing\ref{metaprogramming}.py}] Metaprogramming with Ace, showing how to construct generic functions from abstract syntax trees.}
+%\label{metaprogramming}
+%\end{codelisting}
+
+%\begin{codelisting}
+%\lstinputlisting[linebackgroundcolor={\btLstHL{5-8}}]{listing3.py}
+%\caption{[\texttt{listing\ref{map}.py}] A generic data-parallel higher-order map function targeting OpenCL.}
+%\label{map}
+%\end{codelisting}
+%\begin{codelisting}[t]
+%\lstinputlisting[commentstyle=\color{mauve}]{listing7.py}
+%\caption{[\texttt{listing\ref{metaprogramming}.py}] Metaprogramming with Ace, showing how to construct generic functions from abstract syntax trees.}
+%\label{metaprogramming}
+%\end{codelisting}
 Lines 3-4 introduce \verb|map|, an Ace function of three arguments that is governed by the \emph{active base} referred to by \verb|clx.base| and targeting the \emph{active target} referred to by \verb|clx.opencl|.
@@
 %Not shown are several additional conveniences, such as delegated kernel sizing and \verb|In| and \verb|Out| constructs that can reduce the size and improve the clarity of this code further; due to a lack of space, the reader is referred to the language documentation.
-
-\section{Extensibility}\label{att}
-The core of Ace consists of about 1500 lines of Python code implementing its primary concepts: generic functions, concrete functions, active types, active bases and active targets.
-The latter three comprise Ace's extension mechanism.
-Extensions provide semantics to, and govern the compilation of, Ace functions, rather than logic in Ace's core.
-%Indeed, the name ``Ace'' might itself be an acronym: it is an ``active compilation environment''.
-
-\subsection{Active Types}\label{atypes}
-Active types are the primary means for extending Ace with new abstractions.
-An active type, as mentioned previously, is an instance of a class implementing the \verb|ace.ActiveType| interface.
-Listing \ref{cplx} shows an example of such a class: the \verb|clx.Cplx| class used in Listing \ref{compscript}, which implements the logic of complex numbers.
-The constructor takes as a parameter any numeric type in \verb|clx| (line \ref{cplx}.2).
-
-\subsubsection{Dispatch Protocol}
-\begin{codelisting}
-\lstinputlisting{listing9.py}
-\caption{\texttt{[}in \texttt{examples/clx.py]} The active type family \texttt{Ptr} implements the semantics of OpenCL pointer types.}
-\label{cplx}
-\end{codelisting}
-In a compiler for a monolithic language, there would be a \emph{syntax-directed} protocol governing typechecking and translation.
-In a compiler written in a functional language, for example, one would declare datatypes that captured all forms of types and expressions, and the typechecker would perform exhaustive case analysis over the expression forms.
-That is, all the semantics are implemented in one place.
-The visitor pattern typically used in object-oriented languages implements essentially the same protocol.
-This does not work for an extensible language because new cases and logic need to be added \emph{modularly}, in a safely composable manner.
-
-Instead of taking a syntax-directed approach, the Ace compiler's typechecking and translation phases take a \emph{type-directed approach}.
-When encountering a compound term (e.g.
-\verb|e[e1]|), the compiler defers control over typechecking and translation to the active type of a designated subexpression (e.g.
-\verb|e|) determined by Ace's fixed \emph{dispatch protocol}.
-Below are examples of the choices made in Ace.
-Due to space constraints, we do not show the full protocol.
-\begin{itemize}
-\item Responsibility over {\bf attribute access} (\texttt{e.attr}), {\bf subscripting} (\texttt{e[e1]}) and \textbf{calls} (\verb|e(e1, ..., en)|) and {\bf unary operations} (e.g.
-\verb|-e|) is handed to the type recursively assigned to \texttt{e}.
-\item Responsibility over {\bf binary operations} (e.g.
-\verb|e1 + e2|) is first handed to the type assigned to the left operand.
-If it indicates a type error, the type assigned to the right operand is handed responsibility, via a different method call.
-{Note that this operates like the corresponding rule in Python's \emph{dynamic} operator overloading mechanism; see Sec.
-\ref{related} for a discussion.}
-\item Responsibility over \textbf{constructor calls} (\verb|[t](e1, ..., en)|), where \verb|t| is a \emph{compile-time Python expression} evaluating to an active type, is handed to that type.
-If \verb|t| evaluates to a \emph{family} of types, like \verb|clx.Cplx|, the active type is first generated via a class method, as discussed below.
-%\item Responsibility over {\bf simple assignment statements}, is handed to the type of the variable on the left (which, as we will see below, is determined by the active base).
-If this type does not provide special assignment semantics, the base must handle it.%Destructuring assignment is also supported by a somewhat more complex protocol.\todo{clarify}
-\end{itemize}
-
-\subsubsection{Typechecking}
-When typechecking a compound expression or statement, the Ace compiler temporarily hands control to the object selected by the dispatch protocol by calling the method \verb|type_|$X$, where $X$ is  the name of the syntactic form, taken from the Python grammar \cite{python} (appended with a suffix in some cases).
-
-For example, if \verb|c| is a complex number, then \verb|c.ni| and \verb|c.i| are its non-imaginary and imaginary components, respectively.
-These expressions are of the form \verb|Attribute|, so the typechecker calls \verb|type_Attribute| (line \ref{cplx}.7).
-This method receives the compilation context, \verb|context|, and the abstract syntax tree of the expression, \verb|node| and must return a type assignment for the node, or raise an \verb|ace.TypeError| if there is an error.
-In this case, a type assignment is possible if the attribute name is either \verb|"ni"| or \verb|"i"|, and an error is raised otherwise (lines \ref{cplx}.8-\ref{cplx}.10).
-We note that error messages are an important and sometimes overlooked facet of {ease-of-use} \cite{marceau2011measuring}.
-A common frustration with using general-purpose abstraction mechanisms to encode an abstraction is that they can produce  verbose and cryptic error messages that reflect the implementation details instead of the semantics.
-Ace supports custom error messages.%Active types permit direct encoding of semantics and customization of error messages.
-
-Complex numbers also support binary arithmetic operations partnered with both other complex numbers and with non-complex numbers, treating them as if their imaginary component is zero.
-The typechecking rules for this logic is implemented on lines \ref{cplx}.17-\ref{cplx}.29.
-Arithmetic operations are usually symmetric, so the dispatch protocol checks the types of both subexpressions for support.
-To ensure that the semantics remain deterministic in the case that both types support the binary operation, Ace asks the left first (via \verb|type_BinOp_left|), asking the right (via \verb|type_BinOp_right|) only if the left indicates an error.
-In either position, our implementation begins by recursively assigning a type to the other operand in the current context via the \verb|context.type| method (line \ref{cplx}.24).
-If supported, it applies the C99 rules for arithmetic operations to determine the resulting type (via \verb|c99_binop_t|, not shown).
-
-Finally, a complex number can be constructed inside an Ace function using Ace's special constructor form: \verb|[clx.Cplx](3,4)| represents $3+4i$, for example.
-The term within the braces is evaluated at \emph{compile-time}.
-Because \verb|clx.Cplx| evaluates not to an active type, but to a class, this form is assigned a type by handing control to the class object via the \emph{class method} \verb|type_New|.
-It operates as expected, extracting the types of the two arguments to construct an appropriate complex number type (lines \ref{cplx}.50-\ref{cplx}.57), raising a type error if the arguments cannot be promoted to a common type according to the rules of C99 or if two arguments were not provided.% (an exercise for the reader: modify this method to also allow a single argument for when the imaginary part is 0).
-
-%Note that in each example, subexpressions are not \emph{automatically} assigned types.
-Types must be requested from the context.
-This is useful because it allows types to reinterpret subexpressions of terms they control, to supporting more complex constructs (e.g.
-pattern matching on functional datatypes, see Sec.
-\ref{datatypes}).
-
-\subsubsection{Translation}
-Once typechecking a method is complete, the compiler enters the translation phase, where terms in the target language are generated from Ace terms.
-Terms in the target language are generated by calling methods of the \emph{active target} governing the function being compiled.
-The translation phase operates similarly to typechecking, using the dispatch protocol to invoke methods named \verb|trans_|$X$.
-These methods have access to the context and node just as during typechecking, as well as the active target (named \verb|target| here).
-
-As seen in Listing \ref{mapout}, we are implementing complex numbers internally using OpenCL vector types, like \verb|int2|.
-Let us look first at \verb|trans_New| on lines \ref{cplx}.54-\ref{cplx}.60, where new complex numbers are translated to vector literals by invoking \verb|target.VecLit|.
-This will ultimately  generate the necessary OpenCL code, as a string, to complete compilation (these strings are not directly manipulated by extensions, however, to avoid problems with, e.g.
-precedence).
-For it to be possible to reason compositionally about the correctness of compilation, all complex numbers must translate to terms in the target language that have a consistent target type.
-The \verb|trans_type| method of the \verb|ace.ActiveType| associates a type in the target language, here a vector type like \verb|int2|, with the active type.
-Ace supports a mode where this \emph{representational consistency} is dynamically checked during compilation (requiring that the active target know how to assign types to terms in the target language, which can be done  for our OpenCL target as of this writing).
-
-The translation methods for attributes (line \ref{cplx}.12) and binary operations  (line \ref{cplx}.31) proceed in a straightforward manner.
-The context provides a method, \verb|trans|, for recursively determining the translation of subexpressions as needed.
-Of note is that the translation methods can assume that typechecking succeeded.
-For example, the implementation of \verb|trans_Attribute| assumes that if \verb|node.attr| is not \verb|'ni'| then it must have been \verb|'i'| on line \ref{cplx}.14, consistent with the implementation of \verb|type_Attribute| above it.
-Typechecking and translation are separated into two methods to emphasize that typechecking is not target-dependent, and to allow for more advanced uses, like type refinements and hypothetical typing judgements, that we do not describe here.
-
-\subsection{Active Bases}\label{abases}
-Each generic function is associated with an active base, which is an object implementing the \verb|ace.ActiveBase| interface.
-The active base specifies the \emph{base semantics} of that function.
-It controls the semantics of statements and expressions that do not have a clear ``primary'' subexpression for the dispatch protocol to defer to.
-A base is handed control over typechecking of statements and expressions in the same way as active types: via \verb|type_|$X$ and \verb|trans_|$X$ methods.
-Each function can have a different base.
-
-Literals are the most prominent form given their semantics by an active base.
-Our example active base, \verb|clx.base|, assigns integer literals the type \verb|clx.int32| while floating point literals have type \verb|clx.double|, consistent with the semantics of OpenCL and C99.
-The \verb|clx.base| object is an instance of \verb|clx.Base| that is pre-constructed for convenience.
-Alternative bases can be generated via different constructor arguments.
-For example, \verb|clx.Base(flit_t=clx.float)| is a base semantics where floating point literals have type \verb|clx.float|.
-This is useful because some OpenCL devices do not support double-precision numbers, or impose a significant performance penalty for their use.
-Indeed, to even use the \verb|double| type in OpenCL, a flag must be toggled by a \verb|#pragma| (The OpenCL target we have implemented inserts this automatically when needed, along with some other flags, and annotations like \verb|kernel|).
-Similarly, for some applications, avoiding accidental overflows is more important than performance.
-Infinite-precision integers can be implemented as an active type (not shown) and the base can be asked to use it for numeric literals.
-In all cases, this occurs by inserting a constructor call form, as described above.% the constructor call form, described above, around integers to support this.
-
-The base also initializes the context and governs the semantics of variables and assignment.
-This can also permit it to allow for the use of ``built-in'' operators that were not explicitly imported.
-The semantics of \verb|get_global_id| and \verb|printf| are provided by \verb|clx.base|.
-In fact, the base provides extended versions of them (the former permits multi-dimensional global IDs, the latter supports typechecking format strings).A base which does not provide them as built-ins (requiring that they be imported explicitly) can be generated by passing the \verb|primitives=False| flag.
-
-\subsection{Active Targets and Cross-Compilation}\label{atargets}
-An active target, which we describe being used above, is an instance of \verb|ace.ActiveTarget|.
-It has two primary roles in Ace: 1) providing an API for code generation to a target language, providing term and type constructors for use during translation (and, optionally, a type system implementation to support representational consistency checks); 2) implementing the interoperability layer to support calling of  generic or concrete function directly from Python, as described in Sec.
-\ref{implicit}.
-
-An active type or base can support multiple active targets by simply examining \verb|target| during translation (e.g.
-by performing capability checks using Python's \verb|hasattr| function).
-Alternatively, the active target can simulate the interface of a different target but generate code differently.
-For example, although our implementation of OpenCL's type system as a collection of active types and an active base relies on a target that supports OpenCL terms and types, our \verb|CUDA| and \verb|C99| targets provide partial support for the same term and type forms, in order to support cross-compilation.
-We will show the latter being used in Sec.
-\ref{examples}.
-
-\subsection{Compositional Reasoning}\label{safety}
-Every statement and expression in an Ace function is governed by exactly one active type, determined by the dispatch protocol, or by the single active base associated with the Ace function.
-The representational consistency check ensures that compilation is successful without requiring that types that abstract over other types (e.g.
-container types) know about their internal implementation details.
-Together, this permits compositional reasoning about the semantics of Ace functions even in the presence of many extensions.
-This stands on contrast to many prior approaches to extensibility, where extensions could insert themselves into the semantics in conflicting ways, making the order of imports matter (see Sec.
-\ref{related}).
+\section{Related Work}\label{related}
+\section{Discussion}\label{discussion}
+
+\section{Extensibility}
 

### 2014-03-23 ace-oopsla14/Ptr.py

@@ -1,0 +1,40 @@
+class Ptr(ace.ActiveType):
+  def __init__(self, addr_space, t):
+    self.addr_space = addr_space
+    self.t = t
+
+  def translate_type(self, context):
+    t_x = self.t.translate_type(context)
+    return target.Ptr(self.addr_space, t_x)
+        
+  def type_Subscript(self, context, node):
+    slice_t = context.type(node.slice)
+    if isinstance(slice_t, Integer):
+      return self.t
+    else:
+      raise ace.TypeError('<error message>', node)
+       
+  def translate_Subscript(self, context, node):
+    value_x = context.translate(node.value)
+    slice_x = context.translate(node.slice)
+    return context.target.Subscript(value_x, slice_x)
+
+  def type_BinOp_left(self, context, node):
+    if isinstance(node.operator, ast.Add):
+      right_t = context.type(node.right)
+      if isinstance(right_t, Integer):
+        return self
+    elif isinstance(node.operator, ast.Sub):
+      right_t = context.type(node.right)
+      if self == right_t:
+        return ptrdiff_t
+    raise ace.TypeError('<error message>', node)
+
+  def translate_BinOp_left(self, context, node):
+    left_x = context.translate(node.left)
+    right_x = context.translate(node.right)
+    return context.target.BinOp(left_x, 
+      node.operator, right_x)
+
+  # type_BinOp_right and translate_BinOp_right 
+  # (not shown) are symmetric

### 2014-03-23 ace-oopsla14/datatypes.py

@@ -1,0 +1,29 @@
+class DT(object):
+  pass
+
+class DT_Empty(DT):
+  pass
+  
+class DT_Leaf(DT):
+  def __init__(self, data):
+    self.data = data
+
+class DT_Node(DT):
+  def __init__(self, data):
+    self.data = data
+
+def depth_gt_2(x):
+  if isinstance(x, DT_Node):
+    if isinstance(x.data[0], DT_Node):
+      r = True
+    elif isinstance(x.data[1], DT_Node):
+      r = True
+  elif True:
+    r = False
+  return r
+
+if __name__ == "__main__":
+  my_lil_tree = DT_Node(DT_Empty(), DT_Empty())
+  my_big_tree = DT_Node(my_lil_tree, my_lil_tree)
+  assert not depth_gt_2(my_lil_tree)
+  assert depth_gt_2(my_big_tree)

### 2014-03-23 ace-oopsla14/datatypes_t.py

@@ -1,0 +1,25 @@
+import ace, ace.python as py, examples.fp as fp
+
+Tree = lambda name, a: fp.Datatype(name, 
+  lambda tree: {
+    'Empty': fp.unit,
+    'Leaf': a,
+    'Node': fp.tuple(tree, tree)
+  })
+
+DT = Tree('DT', py.dyn)
+
+@ace.fn(py.Base(trailing_return=True))[[DT]]
+def depth_gt_2(x):
+  x.case({
+    DT.Node(DT.Node(_), _): True,
+    DT.Node(_, DT.Node(_)): True,
+    _: False
+  })
+
+@ace.fn(py.Base(main=True))[[]]
+def __main__():
+  my_lil_tree = DT.Node(DT.Empty, DT.Empty)
+  my_big_tree = DT.Node(my_lil_tree, my_lil_tree)
+  assert not depth_gt_2(my_lil_tree)
+  assert depth_gt_2(my_big_tree)

### 2014-03-23 ace-oopsla14/hello.py

@@ -1,0 +1,11 @@
+from ace.OpenCL import OpenCL, printf
+ 
+print "Hello, compile-time world!"
+ 
+@OpenCL.fn
+def main():
+	hello = "Hello, run-time world!"
+    printf(hello)
+main = main.compile()
+
+print "Goodbye, compile-time world!"

### 2014-03-23 ace-oopsla14/listing11.py

@@ -1,0 +1,11 @@
+class ReduceLIF(ModelNode):
+  """Base class for spiking neuron models."""
+  def in_model_code(self, g): # g is the code generator
+    "idx_state = idx_model + count*" << g
+    "(realization_n - realization_start)" << g
+    self.insert_hook("read_incoming", g)
+    self.insert_hook("read_state", g)
+    self.insert_hook("calculate_inputs", g)
+    self.insert_hook("state_calculations", g)
+    self.insert_hook("spike_processing", g)
+  # ...

### 2014-03-23 ace-oopsla14/listing3.py

@@ -1,0 +1,10 @@
+import ace, examples.clx as clx
+
+@ace.fn(clx.base, clx.opencl)
+def map(input, output, f):
+    thread_idx = get_global_id()
+    output[thread_idx] = f(input[thread_idx])
+    if thread_idx == 0:
+    	printf("Hello, run-time world!")
+
+print "Hello, compile-time world!"

### 2014-03-23 ace-oopsla14/listing4.py

@@ -1,0 +1,10 @@
+import listing1, listing2, ace, examples.clx as clx
+
+T1 = clx.Ptr(clx.global_, clx.float)
+T2 = clx.Ptr(clx.global_, clx.Cplx(clx.int))
+TF = listing2.negate.ace_type
+
+try: map_neg_f32 = listing1.map[[TF, T1, T1]]
+except ace.TypeError as e: print e.full_msg
+map_neg_f32 = listing1.map[[T1, T1, TF]]
+map_neg_ci32 = listing1.map[[T2, T2, TF]]

### 2014-03-23 ace-oopsla14/listing6.py

@@ -1,0 +1,14 @@
+from ace.OpenCL import OpenCL, int, double, long
+
+@OpenCL.fn
+def threshold_scale(x, scale):
+    if x <= 0:
+        y = 0
+    else:
+        y = scale * x
+    return y
+
+f = threshold_scale.compile(int, double)
+g = threshold_scale.compile(int, long)
+assert f.return_type == double
+assert g.return_type == long

### 2014-03-23 ace-oopsla14/listing7.py

@@ -1,0 +1,9 @@
+import ace, examples.clx as clx, ast, astx
+
+_fn = ace.fn(clx.base, clx.opencl)
+
+scale = _fn(ast.parse("""def scale(x, s): 
+	return x * s""")
+
+negate = _fn(astx.specialize(scale.ast, "negate", 
+	s=astx.parse_exp("-1"))

### 2014-03-23 ace-oopsla14/listing8.py

@@ -1,0 +1,12 @@
+import listing1, listing2, examples.clx as clx, numpy
+
+clx.opencl.ctx = clx.Context.for_device(0, 0)
+
+input = numpy.ones((1024,1024))
+d_input = clx.to_device(input)
+d_output = clx.alloc(like=input)
+
+listing1.map(d_input, d_output, listing2.negate, 
+  global_size=d_in.shape)
+
+assert (cl.from_device(d_out) == input * -1).all()

### 2014-03-23 ace-oopsla14/listing9.py

@@ -1,0 +1,63 @@
+class Cplx(ace.ActiveType):
+  def __init__(self, t):
+    if not isinstance(t, Numeric):
+      raise ace.InvalidTypeError("<error message>")
+    self.t = t
+
+  def type_Attribute(self, context, node):
+    if node.attr == 'ni' or node.attr == 'i':
+      return self.t
+    raise ace.TypeError("<error message>", node)
+
+  def trans_Attribute(self, context, target, node):
+    value_x = context.trans(node.value)
+    a = 's0' if node.attr == 'ni' else 's1'
+    return target.Attribute(value_x, a)
+
+  def type_BinOp_left(self, context, node):
+    return self._type_BinOp(context, node.right)
+
+  def type_BinOp_right(self, context, node):
+    return self._type_BinOp(context, node.left)
+
+  def _type_BinOp(self, context, other):
+    other_t = context.type(other)
+    if isinstance(other_t, Numeric):
+      return Cplx(c99_binop_t(self.t, other_t))
+    elif isinstance(other_t, Cplx):
+      return Cplx(c99_binop_t(self.t, other.t))
+    raise ace.TypeError("<error message>", other)
+
+  def trans_BinOp(self, context, target, node):
+    r_t = context.type(node.right)
+    l_x = context.trans(node.left)
+    r_x = context.trans(node.right)
+    make = lambda a, b: target.VecLit(
+      self.trans_type(self, target), a, b)
+    binop = lambda a, b: target.BinOp(
+      a, node.operator, b)
+    si = lambda a, i: target.Attribute(a, 's'+str(i))
+    if isinstance(r_t, Numeric):
+      return make(binop(si(l_x, 0), r_x), si(r_x, 1))
+    elif isinstance(r_t, Cplx):
+      return make(binop(si(l_x, 0), si(r_x, 0)),
+        binop(si(l_x, 1), si(r_x, 1)))
+
+  @classmethod
+  def type_New(cls, context, node):
+    if len(node.args) == 2:
+      t0 = context.type(node.args[0])
+      t1 = context.type(node.args[1])
+      return cls(c99_promoted_t(t0, t1))
+    raise ace.TypeError("<error message>", node)
+
+  @classmethod
+  def trans_New(cls, context, target, node):
+    cplx_t = context.type(node)
+    x0 = context.trans(node.args[0])
+    x1 = context.trans(node.args[1])
+    return target.VecLit(cplx_t.trans_type(target), 
[…]

### 2014-03-23 ace-oopsla14/ooclxfp.py

@@ -1,0 +1,10 @@
+import examples.clx as clx, examples.fp as fp, 
+  examples.oo as oo
+
+A = clx.Struct('A', {'x': clx.int})
+B = fp.Record('B', {'y': clx.float})
+C = oo.Prototypic('C', A, B)
+
+@ace.fn(clx.base, clx.c99)
+def test(input : C):
+  return input.x + input.y

### 2014-03-23 ace-oopsla14/ace-oopsla14.tex
Finished draft of oopsla14 through sec 2 with new examples

@@
 
-\subsection{Introductory Forms}
+\subsection{Bidirectional Typechecking of Introductory Forms}
 On lines 21-30, we define the function \verb|__toplevel__| (the base will consider this a special name for the purposes of compilation, discussed in the Sec.
@@
 After printing a run-time greeting, we construct a value of type \verb|C| on lines 24-26.
-Recall that \verb|C| is a record type, so we provide the names of the fields (here, without quotes because we are no longer in the type-level langauge) and values of the appropriate type using syntax for a dictionary literal.
-Because there may be multiple types that use this form, we say which type the literal should be \emph{analyzed} against with the \emph{literal ascription}  \verb|(C)|.
-This is another repurposing of Python's syntax: here function application.
-When the ``function'' has one of Python's literal forms and the (single) argument is a type, we treat it instead as an \emph{ascribed introductory form}.
-The literal is treated as an introductory form for that type (we will see precisely how this occurs shortly).
+Recall that \verb|C| is a record type, so we provide the names of the fields (here, without quotes because we are no longer in the type-level language) and values of the appropriate type using syntax for a dictionary literal.
+We must specify which type the literal should be {analyzed} against by giving a \emph{literal ascription}  \verb|(C)|.
+This again repurposes existing  syntax: here function application.
+When the ``function'' is of literal form (dictionaries, tuples, lists, strings, numbers and keywords like \verb|True| and \verb|None|) and the argument is a type, we consider it instead an \emph{ascribed introductory form} of that type  (we will see precisely how it is checked shortly).
 We see the same form used with tuple literals on lines 27-28 to introduce terms of type \verb|Transfer|.
 
-\subsection{Bidirectional Typechecking}
+%\subsection{Bidirectional Typechecking}
 The string, number and dictionary forms inside the outermost forms on lines 24-28 are also introductory forms, but we do not include an ascription because the definition of \verb|C| determines which types they should be analyzed against.
 As we will discuss shortly, Ace has a \emph{bidirectional type system} distinguishing locations where a type must be \emph{synthesized} from an expression (e.g.
-at the top-level) and locations where an expression is \emph{analyzed} against a type \cite{LovasPfenning08}.
-The base determines the type a literal should be analyzed against in a synthetic context.
-For example, if we did not specify the literal ascription \verb|(C)| on line 26, the base would analyze it against the Python dictionary type, which would cause a type error because the variables used as keys were not defined in the surrounding context.
-
-\subsection{Compilation}\label{compilation}
+at the top-level) and locations where the expression must be \emph{analyzed} against a known type \cite{LovasPfenning08}.
+Literal forms can only be analyzed against a type.
+For example, if they are used in a synthetic context, the base determines a ``default ascription''.
+If we had not specified an ascription on line 26, the literal would need to synthesize a type, so the base, \verb|py|, would analyze it against the Python dictionary type, which would cause a type error because the variables used as keys are not bound.
+
+An ascription can also be a {type constructor}, rather than a type.
+For example, we might write \verb|[1, 2] (matrix)| rather than \verb|[1, 2] (matrix[int])| assuming the default ascription for integer literals is \verb|int|.
+The type constructor synthesizes an index based on the types synthesized by the subexpressions.
+If we wanted a matrix of \verb|float|s, then we would have to write either \verb|[1, 2] (matrix[float])| or \verb|[1(float), 2(float)] (matrix)|.
+We will see how this works in greater technical detail shortly.
+
+\subsection{External Compilation}\label{compilation} 
+To typecheck, translate and execute the code in Listing \ref{example}, we have two choices: do so externally at the shell, or perform compilation interactively from within Python.
+We will first describe the former, then return to interactive compilation in Sec.
+\ref{targets} with a different example.
+Listing \ref{external-compilation} shows how to use the \verb|acec| compiler to typecheck and translate \texttt{listing\ref{example}.py}, producing a file \verb|_listing1.py| which can then be executed by the standard Python interpreter.
+Note that the \textbf{\texttt{print}} statement at the top level of the compilation script was indeed evaluated during compilation.
+These two steps can be combined by running \verb|ace listing1.py| (the intermediate file is not generated in this case unless requested).
+
+The \verb|acec| compiler (a simple Python script) operates in two stages:
+\begin{enumerate}
+\item Executes the provided compilation script (\verb|listing1.py|)
+\item For any top-level bindings that are Ace functions in the final environment (with dynamic type \verb|ace.ConcreteFn|, as we will discuss), it initiates the active type-checking and translation process described in Sec.
+\ref{att}.
+If no type errors are discovered, the translations are extracted and emitted as files (their extensions are determined by the targets in use, discussed in Sec.
+\ref{targets}; here our target is  the default target associated with the base \verb|examples.py.py|).
+If a type error is discovered, no file is emitted and the error message is displayed on the console.
+\end{enumerate}
+
+In our example, there are no type errors, so the file \verb|_listing1.py| is generated (Listing \ref{example-out}). Notice that:
+\begin{enumerate}
+\item The base placed the translation of the body of the function \verb|__toplevel__| at the top level of the resulting file.
+\item Records translated to tuples (or, if there was only a single field, its value was passed around unadorned). The field names were needed only statically.
+\item Terms of type \verb|string_in[r"..."]| translated to strings.
+Checks for well-formedness were rendered unnecessary statically by the type system.\todo{Nathan?}\footnote{Note that there are situations (e.g.
+if a string is read in from the console) that would necessitate an initial run-time check, but no further checks in downstream functions would be necessary.
+See the appendix for details.}
+\item Decimals are represented as pairs of integers.
+Conversion to a string happens via a helper function defined in a ``runtime'' package imported with an internal name, \verb|__ace_0|, to avoid naming conflicts.
+\item Prototypic objects are represented as pairs consisting of the fore and the prototype. Dispatch to the appropriate record based on the field name was static (line 5).
+\end{enumerate}
+
+The file \verb|_listing1.py| is meant only to be executed.
+The invariants necessary to ensure that execution does not ``go wrong'' were checked statically and entities having no bearing on execution, like field names and types themselves, were erased.
+As a result, execution is essentially as fast as it could be given the target language chosen (Python).
+We will target an even faster language, \verb|OpenCL|, in Sec.
+\ref{targets}.
 
 %Because \verb|t| will be statically guaranteed to have type \verb|Transfer|,  
-\subsection{B}
+\subsection{Type Errors}
+Listing \ref{oops} shows an example of code containing several type errors.
+Indeed, lines 9-13 each contain a type error.
+More alarmingly, if this code were written in a dynamically-typed language, they would only be found if the code was run on the first day of the month.
+Static type checking allows us to find these errors during compilation, without needing to execute the code.
+Listing \ref{oops-compilation} shows the result of attempting to compile this code.
+The compilation script completes (so that functions can refer to each other mutually recursively), then the Ace functions are typechecked.
+The typechecker raises an exception at the first error, and \verb|acec| prints it.
+\begin{codelisting}[t]
+\begin{lstlisting}
+from listing1 import A, C, log_transfer
+from datetime import date
+
+@py
+def pay_oopsie(a):
+  {a : A}
+  print "Checking date..."
+  if date.today().day == 1:
+    common = {nome: "Oopsie Daisy",
+              account_num: None,
+              routing_num: "0-0000-0001"} (C)
+    log_transfer((common, a))
+    a.amount += 1000
+
+print date.today().day
+\end{lstlisting}
+\caption{[\texttt{listing\ref{oops}.py}] Lines 9-13 each have type errors.}
+\label{oops}
+\end{codelisting}
+\begin{codelisting}
+\begin{lstlisting}[style=Bash]
+$ `acec listing4.py`
+2
+[acec] TypeError in listing4.py (line 8, col 15): 
+  [record] Invalid field name: nome
+           Expected: name, account_num, routing_num
+\end{lstlisting}
+\caption{Compiling \texttt{listing\ref{oops}.py} using \texttt{acec} catches the errors statically (compilation stops at first error).}
+\label{oops-compilation}
+\end{codelisting}
+
 \section{Active Type Synthesis and Translation}\label{att}
-The core of Ace consists of about 1500 lines of Python code implementing its primary concepts: generic functions, concrete functions, active types, active bases and active targets.
+\todo{To be revised from here on down}The core of Ace consists of about 1500 lines of Python code implementing its primary concepts: generic functions, concrete functions, active types, active bases and active targets.
 The latter three comprise Ace's extension mechanism.
@@
 To produce an output file from an Ace ``compilation script'' like \verb|listing|\texttt{\ref{compscript}}\verb|.py|, the command \verb|acec| can be invoked from the shell, as shown in Listing \ref{mapc}.
-The \verb|acec| compiler (a simple Python script) operates in two stages:
-\begin{enumerate}
-\item Executes the provided Python file (\verb|listing3.py|).
-\item Extracts the translations from concrete functions and other top-level constructs (e.g.
-types requiring declarations, or generated imports and pragmas) in the top-level Python environment.
-This may produce one or more files as mediated by the active targets that were used (here, just \verb|listing3.cl|, but a web framework built upon Ace might produce separate HTML, CSS and JavaScript files; see Sec.
-\ref{atargets}).
-\end{enumerate}
-
-In this case, stage 1 results in the output on lines \ref{mapc}.2-\ref{mapc}.4.
-The type error printed on lines \ref{mapc}.3-\ref{mapc}.4 will be explained in the next section.
-The compiler then enters stage 2 and concludes with the message on line \ref{mapc}.5 to indicate that one file was generated.
-This file is shown in Listing \ref{mapout} and can be used by any programs that consume OpenCL code (e.g.
-a C program that invokes the generated kernels via the OpenCL host API).
-We will show in Section \ref{backend} that for targets with Python bindings, such as OpenCL, CUDA, C, Java or Python itself, generic functions can be executed directly, without any of the explicit compilation steps in Listings \ref{compscript}-\ref{mapc}.
-
 \subsection{Types}
@@
 %Not shown are several additional conveniences, such as delegated kernel sizing and \verb|In| and \verb|Out| constructs that can reduce the size and improve the clarity of this code further; due to a lack of space, the reader is referred to the language documentation.
-\section{Related Work}\label{related}
-\section{Discussion}\label{discussion}
-
-\section{Extensibility}
 

### 2014-03-24 ace-oopsla14/ace-oopsla14.tex
OOPSLA14 progress

@@
 Programmers are justifiably reluctant to adopt new language dialects to access stronger type systems.
-This suggests a need for a language that is \emph{compatible} with existing libraries, tools and infrastructure and that has an \emph{internally extensible type system}, so that adopting and combining type systems requires only importing libraries in the usual way, without the need to resolve link-time ambiguities.
+This suggests a need for a language that is \emph{compatible} with existing libraries, tools and infrastructure and that has an \emph{internally extensible type system}, so that adopting and combining type systems requires only importing libraries in the usual way, without the possibility of link-time ambiguities or safety issues.
 
@@
 Python serves as Ace's type-level language.
-Rather than building in a particular type system, the Ace compiler inverts control over type assignment and translation to type-level functions according to a type-directed protocol that does not lead to ambiguities when type systems are combined.
-We show that this protocol is flexible enough to admit library-based implementations of types  drawn from functional, object-oriented, parallel, low-level and domain-specific languages.
-We also show how this permits the construction of safe, natural and extensible foreign function interfaces using a novel form of ``staged'' type propagation from Python into Ace.
-We also give a simplified calculus that describes how a type-level language can be used to enable type system extensions and demonstrates how type safety can be maintained by lifting a technique from the typed compilation literature into the language.%To show that the method scales, we give a full implementation of the OpenCL programming language as a library.
+Rather than building in a particular type system, Ace has an extensible bidirectional type system that inverts control over typing and translation to type-level functions according to a protocol that cannot lead to ambiguities and type safety issues when type systems are combined.
+We give a simplified calculus that describes the type theoretic foundations of this mechanism and show that, as implemented in Ace, it is flexible enough to admit practical library-based implementations of types drawn from functional, object-oriented, parallel and domain-specific languages.
+We also show how to construct safe, natural and extensible foreign function interfaces using a novel form of ``staged'' type propagation from Python into Ace.
+% can be used to enable type system extensions and demonstrates how type safety can be maintained by lifting a technique from the typed compilation literature into the language.%To show that the method scales, we give a full implementation of the OpenCL programming language as a library.
 %We briefly examine the type theoretic foundations of this approach and give modular criteria that guarantee that an extension is safe and modular.
@@
 
-Unfortunately, researchers and domain experts (\emph{providers}) designing potentially useful new abstractions can sometimes find it difficult to implement them as libraries, particularly when they require strengthening a language's type system.
+Unfortunately, researchers and domain experts aiming to provide potentially useful new abstractions can sometimes find it difficult to implement them as libraries, particularly when they require strengthening a language's type system.
 In these situations, abstraction providers often develop a new language or language dialect.
-Unfortunately, this \emph{language-oriented approach} \cite{language-oriented} does not scale to large applications: using  components written in languages with different type systems can be awkward and lead to safety problems and performance overhead at interface boundaries.
+Unfortunately, this \emph{language-oriented approach} \cite{journals/stp/Ward94} does not scale to large applications: using  components written in languages with different type systems can be awkward and lead to safety problems and performance overhead at interface boundaries.
 %To avoid this, it is necessary to combine the different languages and their associated tools, but this often requires significant effort and expertise (we will discuss this further in Sec.
@@
 Unfortunately, the mechanisms available today have several problems.
-First, they are themselves often available only within a dialect of an existing language and thus face a ``chicken-and-egg'' problem: a language like SugarJ \cite{sugarj} must overcome the same extrinsic issues as a language like HJ.
+First, they are themselves often available only within a dialect of an existing language and thus face a ``chicken-and-egg'' problem: a language like SugarJ \cite{erdweg2011sugarj} must overcome the same extrinsic issues as a language like HJ.
 Second, giving providers too much control over the language can lead to intrinsic safety issues, ambiguities and conflicts between extensions, as we will discuss.
@@
 
-This paper describes the design and implementation of Ace, an internally extensible, statically typed programming language designed considering both these extrinsic and intrinsic factors.
-To address the ``chicken-and-egg'' problem, Ace is implemented within Python, entirely as a library  \cite{Politz:2013:PFM:2509136.2509536,python}.
-The top-level of an Ace file can be thought of as a \emph{compilation script} written in Python and, as we will discuss, Python serves as Ace's type-level language.
-Ace and Python share a common syntax and package system, so users of Ace can leverage established tools and infrastructure directly.
+This paper describes the design and implementation of Ace, an internally extensible, statically typed programming language designed considering both extrinsic and intrinsic factors.
+To address the ``chicken-and-egg'' problem, Ace is implemented within Python \cite{Politz:2013:PFM:2509136.2509536,python}, entirely as a library.
+The top-level of an Ace file is a \emph{compilation script} written in Python and Python serves as Ace's \emph{type-level language}, as we will discuss.
+Ace and Python share a common syntax and package system, so users of Ace can directly leverage established tools, infrastructure and coding standards.
 %The semantics of  define run-time behavior.
 
-Functions marked by a decorator as under the control of Ace are \emph{statically typechecked}.
-This makes it possible to rule out many potential issues ahead of execution and thus avoid the need for costly dynamic checks.
+Functions marked by a decorator as under the control of Ace are, unlike Python functions, \emph{statically} typechecked.
+This makes it possible to rule out many potential issues ahead of execution, and thus  avoid  costly dynamic checks.
 % Instead, Ace has a statically-typed semantics that can be extended by users from within libraries.
 However, there is \emph{no particular type system built into Ace}.
-Instead, a type is any  {compile-time object} implementing the \verb|ace.Type| interface.
+Instead, a type is any  object implementing the \verb|ace.Type| interface constructed by the compilation script.
 We call these \emph{active types}.
@@
 to the type of \verb|e|, to determine how to assign a type to the expression as a whole.
-As we will show, this {type-directed protocol}, which we call \emph{active type assignment}, permits the expression of a rich variety of type systems as libraries.
-Unlike syntax-directed protocols, there cannot be ambiguities when separately defined type sytems are combined (extensions have exclusive control over non-overlapping sets of expressions by construction).
+As we will show, this {type-directed protocol}, \emph{active typing}, permits the expression of a rich variety of type systems as libraries.
+Unlike syntax-directed protocols, there cannot be ambiguities when separately defined type sytems are combined (types control non-overlapping sets of expressions by construction).
 Base cases, e.g.
-variables and literals, are handled on a {per-function basis} by a \emph{base semantics}, also a compile-time object implementing an interface, \verb|ace.Base|.
+variables and certain statement forms, are handled on a {per-function basis} by a \emph{base semantics}, also a compile-time object implementing an interface, \verb|ace.Base|.
 
@@
 %One base or type can support multiple targets. 
-When targeting a typed language, care must be taken to ensure that well-typed Ace expressions have well-typed translations, so we integrate a technique for compositionally reasoning about compiler correctness (developed for the TIL compiler for Standard ML \cite{TIL}) into the language.
+When targeting a typed language, care must be taken to ensure that well-typed Ace expressions have well-typed translations, so we integrate a technique for compositionally reasoning about compiler correctness (developed for the TIL compiler for Standard ML \cite{tarditi+:til-OLD}) into the language.
 
@@
 Ace functions targeting a language with Python bindings (e.g.
-Python itself, as well as the others just described) can also be compiled and invoked interactively from Python scripts.
+Python itself, as well as the others just described) can also be compiled and invoked interactively from Python.
 Ace supports a form of \emph{ad hoc polymorphism} (based on singleton types) whereby types propagate into generic functions.
@@
 The remainder of the paper is organized as follows: in Sec.
-\ref{usage}, we describe the basic structure and usage of Ace with an example that simultaneously uses dynamically typed terms (which can be thought of as having the static type \verb|dyn|) alongside statically typed immutable records and a simple prototypic object system, all implemented orthogonally as active types.
+\ref{usage}, we describe the basic structure and usage of Ace with an example that uses several interesting Ace type libraries.
 In Sec.
-\ref{att}, we detail how active type assignment and translation delegate control, at compile-time, to active types and bases.
+\ref{att}, we detail how active typing and translation delegate control, at compile-time, to active types and bases, and discuss how this represents a form of extensibility that side-steps the expression problem.
 In Sec.
-\ref{theory}, we reduce this mechanism to  a core lambda calculus, $\lamAce$,  clarify the relationship to   type-level computation and show how type safety is maintained.
+\ref{theory}, we reduce this mechanism to  a core lambda calculus, $\lamAce$,  clarify the relationship to   type-level computation and show how type safety can be maintained.
 In Sec.
@@
 \ref{discussion} by summarizing our contributions, highlighting the key features needed by a host language to support these mechanisms, and discussing limitations and potential future work.
-The appendix contains details omitted from the main body and additional examples.
+The appendix contains details omitted from the main body and several additional examples.
 
@@
 Ace requires no modifications to the language or features specific to the primary implementation, CPython (so Ace supports alternative implementations like Jython and PyPy).
-This choice pays immediate dividends on the first five lines: Ace's package system is Python's package system, so Python's build tools (e.g.
+This choice pays immediate dividends on the first five lines: Ace's import mechanism is Python's import mechanism, so Python's build tools (e.g.
 \verb|pip|) and package repostories (e.g.
@@
 datatypes, classes, structs) can only be declared.
-Put another way, Ace supports \emph{type-level computation} \cite{tlc} and Python is its type-level language (discussed further in Sec.
-\ref{theory}).
+Put another way, Ace supports \emph{type-level computation} and Python is its type-level language (discussed further in Sec.
+\ref{theory} and Sec.
+\ref{related}).
 %The compilation script can assign shorter names to types for convenience (removing the need for  facilities like \verb|typedef| in C or \verb|type| in Haskell). 
@@
 The field \verb|name| has type \verb|string|, defined in \verb|examples.py|, while the fields \verb|account_num| and \verb|routing_num| have more interesting types, classifying strings guaranteed statically to be in the language of a regular expression  (written, to avoid needing to escape backslashes, using Python's \emph{raw string literals}).
-Our appendix specifies the typechecking rules, based on \cite{nfulton-thesis}.
-We include it to emphasize that we aim to support specialized type systems,  not just general purpose constructs like records and numbers.
-\item Prototypic inheritance is a feature of some dynamically-typed languages (most notably, Self \cite{self} and Javascript).
-On line 12, we consider a simple statically-typed variant.
+Our appendix specifies the typechecking rules, based on \cite{fulton-thesis}.
+We include it to emphasize that we aim to support specialized type systems,  not just general purpose constructs like records, numbers and strings.
+\item 
+%Prototypic inheritance  is a feature of some dynamically-typed languages (most notably, Self \cite{Ungar:Smith:oopsla:1987} and Javascript). 
+On line 12, we consider a simple \emph{prototypic object type} \cite{Lie86}.
 The type \verb|Transfer| classifies terms consisting of a \todo{what is the right term for this?}\emph{fore} of type \verb|A| and a \emph{prototype} of type \verb|C|.
 If a field cannot be found in the fore, the type system will delegate (here, statically) to the prototype.
-This makes it easy to share the values of the fields of a common prototype amongst many fores.
-In our example, we will perform multiple transfers differing only in amount.
+This makes it easy to share the values of the fields of a common prototype amongst many fores, here to allow us to describe multiple transfers differing only in amount.
 \end{enumerate}
 \subsection{Functions}
-Functions implementing run-time behavior are distinguished by a decorator specifying their  \emph{base semantics} (here \verb|py| from \verb|examples.py|; we will discuss this further below).
+Functions implementing run-time behavior (\emph{typed functions}) are distinguished by the presence of a decorator specifying their  \emph{base semantics} (here \verb|py| from \verb|examples.py|; we will discuss this further below).
 These functions are, however, still written using Python's syntax, a choice that is again valuable for extrinsic reasons: users of Ace can use a variety of  tools designed to work with Python source code  without  modification, including code highlighters (like the one used in generating this paper), editor plugins, style checkers and documentation generators.
-We will see how, with a bit of cleverness, Python's syntax can support a variety of static semantics.
+We will see how, with a bit of cleverness, Python's syntax can be repurposed within these functions to support a variety of static and dynamic semantics that differ from Python's.
+Ace  leverages the \verb|inspect| and \verb|ast| modules in the Python standard library to extract abstract syntax trees for typed functions \cite{python}.
 \begin{codelisting}
@@
 On lines 14-18, we see the function \verb|log_transfer|.
-As with a Python function, we write a documentation string like the one shown specifying, informally, the behavior of the function.
-Before moving into the body, however, we  also write a \emph{type signature} (line 17) stating simply that the type of \verb|t| is \verb|Transfer|, the prototypic object type described above.
-As a result, we can assume that \verb|t.amount| and \verb|t.name| have types \verb|decimal[2]| and \verb|string|, respectively.
+As in Python, we write a documentation string  that informally specifies the behavior of the function.
+Before moving into the body, however, we  also write a \emph{type signature} (line 17) stating that the type of \verb|t| is \verb|Transfer|, the prototypic object type described above.
+As a result, we can assume on line 19 that \verb|t.amount| and \verb|t.name| have types \verb|decimal[2]| and \verb|string|, respectively.
 We will return to the details of \verb|print| and the form \verb|[string](t.amount)|, which performs an explicit conversion, shortly.\todo{when?}
 
-Line 17 borrows Python's syntax for dictionary literals to approximate  conventional notation for type annotations.\footnote{If the right-hand side of every annotation is not a type, it will be an error.
-In the rare situation where there are no type annotations and the first statement is a dictionary literal, users can use Python's \texttt{pass} keyword to skip the line.} % (and following conventions for documentation strings, which also describe ``types'' at the end).
+Line 17 borrows Python's syntax for dictionary literals to approximate  conventional notation for type annotations.\footnote{In the rare situation where there are no type annotations and the first statement is a dictionary literal, users can use Python's \texttt{pass} keyword.} % (and following conventions for documentation strings, which also describe ``types'' at the end).
 % The ``keyword'' \verb|type| was chosen because it is the name of a built-in function in Python (and thus would not be used in the way shown for any other purpose).
- {In version 3.0 of Python, syntax for annotating arguments with arbitrary metadata using \verb|:| was introduced,  intended for use by future tools (further suggesting that users of dynamically typed languages see potential in a static typing discipline)  \cite{PEPwhatever}.
-Ace supports both notations when the compilation script is run using Python 3.\textit{x}, but we use the more universally available notation here to satisfy our extrinsic goals: the Python 2.\text{x} series  remains the most widely used by a large margin as of this writing (we will consider cross-compilation in Sec.
-\ref{targets}, however).
+ {In version 3.0 of Python, syntax for annotating arguments with arbitrary values directly was introduced \cite{PEP}:
+% \vspace{-5px}
+\begin{lstlisting}[numbers=none]
+def log_transfer(t : Transfer): ...
+\end{lstlisting}
+\vspace{-5px}
+These annotations were initially intended to serve only as documentation (suggesting that users of dynamically typed languages see potential in a typing discipline, if not a static one), but they were also made available as metadata for use by unspecified  future libraries.
+Ace supports both notations when the compilation script is run using Python 3.\textit{x}, but we use the more universally available notation here in view of our extrinsic goals: the Python 2.\text{x} series  remains the most widely used by a large margin as of this writing (though we consider cross-compilation in Sec.
+\ref{targets}).
 
@@
 \ref{compilation}).
-After printing a run-time greeting, we construct a value of type \verb|C| on lines 24-26.
-Recall that \verb|C| is a record type, so we provide the names of the fields (here, without quotes because we are no longer in the type-level language) and values of the appropriate type using syntax for a dictionary literal.
-We must specify which type the literal should be {analyzed} against by giving a \emph{literal ascription}  \verb|(C)|.
+After printing a run-time greeting, we construct a \emph{value} of type \verb|C| on lines 24-26.
+Recall that \verb|C| is a record type, so we provide the names of the fields (here, without quotes because we are no longer in the type-level language) and values of the appropriate type using the syntactic form normally used for dictionary literals.
+We specify which record type the literal should be {analyzed} against by giving a \emph{literal ascription}, \verb|(C)|.
 This again repurposes existing  syntax: here function application.
-When the ``function'' is of literal form (dictionaries, tuples, lists, strings, numbers and keywords like \verb|True| and \verb|None|) and the argument is a type, we consider it instead an \emph{ascribed introductory form} of that type  (we will see precisely how it is checked shortly).
-We see the same form used with tuple literals on lines 27-28 to introduce terms of type \verb|Transfer|.
+When the ``function'' is of literal form (dictionaries, tuples, lists, strings, numbers and keywords like \verb|True| and \verb|None|) and the argument is a type, we consider it instead an \emph{ascribed introductory form} of that type.
+We see the same form used with Python's tuple syntax on lines 27-28 to introduce terms of type \verb|Transfer| by providing the fore and prototype.
 
 %\subsection{Bidirectional Typechecking}
-The string, number and dictionary forms inside the outermost forms on lines 24-28 are also introductory forms, but we do not include an ascription because the definition of \verb|C| determines which types they should be analyzed against.
-As we will discuss shortly, Ace has a \emph{bidirectional type system} distinguishing locations where a type must be \emph{synthesized} from an expression (e.g.
-at the top-level) and locations where the expression must be \emph{analyzed} against a known type \cite{LovasPfenning08}.
+The string, number and dictionary forms inside the outermost forms on lines 24-28 are also introductory forms, but do not require an ascription.
+This is because the definition of the type \verb|C| already determines which types they should be analyzed.
+As we will discuss in more detail shortly, Ace is built around a \emph{bidirectional type system} that distinguishes locations where an expression must \emph{synthesize} a type (e.g.
+to the right of bindings) from those where it can be \emph{analyzed} against a known type (e.g.
+as an argument to a function with known type) \cite{Lovas08abidirectional}.
 Literal forms can only be analyzed against a type.
-For example, if they are used in a synthetic context, the base determines a ``default ascription''.
-If we had not specified an ascription on line 26, the literal would need to synthesize a type, so the base, \verb|py|, would analyze it against the Python dictionary type, which would cause a type error because the variables used as keys are not bound.
+If used in a synthetic location, the base must specify a ``default ascription''.
+For example, if we had not specified the literal ascription on line 26, the base would cause it to be analyzed against a type for Python dictionaries, which would lead to a type error because the variables used as keys are not bound in the surrounding context.
 
 An ascription can also be a {type constructor}, rather than a type.
-For example, we might write \verb|[1, 2] (matrix)| rather than \verb|[1, 2] (matrix[int])| assuming the default ascription for integer literals is \verb|int|.
+For example, we might write \verb|[1, 2] (matrix)| rather than \verb|[1, 2] (matrix[i32])| assuming the default ascription for integer literals is \verb|i32|.
 The type constructor synthesizes an index based on the types synthesized by the subexpressions.
-If we wanted a matrix of \verb|float|s, then we would have to write either \verb|[1, 2] (matrix[float])| or \verb|[1(float), 2(float)] (matrix)|.
+If we wanted a matrix of \verb|f32|s, then we would have to write either \verb|[1, 2] (matrix[f32])| or \verb|[1(f32), 2(f32)] (matrix)|, or use a base with a different default ascription.
 We will see how this works in greater technical detail shortly.
@@
 To typecheck, translate and execute the code in Listing \ref{example}, we have two choices: do so externally at the shell, or perform compilation interactively from within Python.
-We will first describe the former, then return to interactive compilation in Sec.
+We will here describe the former, then return to interactive compilation in Sec.
 \ref{targets} with a different example.
-Listing \ref{external-compilation} shows how to use the \verb|acec| compiler to typecheck and translate \texttt{listing\ref{example}.py}, producing a file \verb|_listing1.py| which can then be executed by the standard Python interpreter.
-Note that the \textbf{\texttt{print}} statement at the top level of the compilation script was indeed evaluated during compilation.
-These two steps can be combined by running \verb|ace listing1.py| (the intermediate file is not generated in this case unless requested).
-
-The \verb|acec| compiler (a simple Python script) operates in two stages:
+
+Listing \ref{external-compilation} shows how to invoke the \verb|acec| compiler at the shell to typecheck and translate \texttt{listing\ref{example}.py}, resulting in a file named \verb|_listing1.py|.
+This is then sent to the  Python interpreter to produce the desired run-time behavior.
+Note that the {\texttt{print}} statement on line 32 of the compilation script was  evaluated during compilation only.
+These two steps can be combined by running \verb|ace listing1.py| (the intermediate file is not generated in this case).
+
+The Ace compiler is itself a Python library, and \verb|acec| is a simple Python script that invokes it, operating in two steps:
 \begin{enumerate}
 \item Executes the provided compilation script (\verb|listing1.py|)
-\item For any top-level bindings that are Ace functions in the final environment (with dynamic type \verb|ace.ConcreteFn|, as we will discuss), it initiates the active type-checking and translation process described in Sec.
+\item For any top-level bindings that are Ace functions in the final environment (instances of \verb|ace.TypedFn|, as we will discuss), it initiates the active type-checking and translation process described in Sec.
 \ref{att}.
-If no type errors are discovered, the translations are extracted and emitted as files (their extensions are determined by the targets in use, discussed in Sec.
-\ref{targets}; here our target is  the default target associated with the base \verb|examples.py.py|).
-If a type error is discovered, no file is emitted and the error message is displayed on the console.
+If no type errors are discovered, the translations are extracted and emitted (the file extension(s) are determined by the target(s) in use, discussed in Sec.
+\ref{targets}; here our target is  the default target associated with the base \verb|examples.py.py|, which emits Python 2.6 files).
+If a type error is discovered, no file is emitted and the error is displayed on the console.
 \end{enumerate}
 
-In our example, there are no type errors, so the file \verb|_listing1.py| is generated (Listing \ref{example-out}). Notice that:
+In our example, there are no type errors, so the file \verb|_listing1.py|, shown in Listing \ref{example-out},  is generated.
+This file is meant only to be executed.
+The invariants necessary to ensure that execution does not ``go wrong'' were checked statically and entities having no bearing on execution, like field names and types themselves, were erased.
+Notice that:
 \begin{enumerate}
-\item The base placed the translation of the body of the function \verb|__toplevel__| at the top level of the resulting file.
-\item Records translated to tuples (or, if there was only a single field, its value was passed around unadorned). The field names were needed only statically.
+\item The base, recognizing the function name \verb|__toplevel__| as special, placed the translation of its body  at the top level of the file.
+Ace does not have any special function names itself; this is a feature of the chosen base.
+\item Records with two or more fields translated to tuples of their values. If there was only a single field, like \verb|A|, its value was passed around unadorned. 
+\item Decimals are represented as pairs of integers.
+Conversion to a string happens via a helper function defined in a ``runtime'' package imported with an internal name, \verb|__ace_0|, to avoid naming conflicts.
 \item Terms of type \verb|string_in[r"..."]| translated to strings.
-Checks for well-formedness were rendered unnecessary statically by the type system.\todo{Nathan?}\footnote{Note that there are situations (e.g.
+Checks for well-formedness were performed entirely statically by the type system.\todo{Nathan?}\footnote{Note that there are situations (e.g.
 if a string is read in from the console) that would necessitate an initial run-time check, but no further checks in downstream functions would be necessary.
 See the appendix for details.}
-\item Decimals are represented as pairs of integers.
-Conversion to a string happens via a helper function defined in a ``runtime'' package imported with an internal name, \verb|__ace_0|, to avoid naming conflicts.
-\item Prototypic objects are represented as pairs consisting of the fore and the prototype. Dispatch to the appropriate record based on the field name was static (line 5).
+\item Prototypic objects are represented as pairs consisting of the fore and the prototype. Dispatch to the appropriate record based on the field name is static (line 5).
 \end{enumerate}
 
-The file \verb|_listing1.py| is meant only to be executed.
-The invariants necessary to ensure that execution does not ``go wrong'' were checked statically and entities having no bearing on execution, like field names and types themselves, were erased.
-As a result, execution is essentially as fast as it could be given the target language chosen (Python).
-We will target an even faster language, \verb|OpenCL|, in Sec.
-\ref{targets}.
+% As a result, execution is essentially as fast as it could be given the target language chosen (Python). We will target an even faster language, \verb|OpenCL|, in Sec. \ref{targets}.
 
@@
 Indeed, lines 9-13 each contain a type error.
-More alarmingly, if this code were written in a dynamically-typed language, they would only be found if the code was run on the first day of the month.
-Static type checking allows us to find these errors during compilation, without needing to execute the code.
+If analagous code were written in Python itself, these could only be found if the code was executed on the first day of the month (and, depending on the implementation, not all of the issues would  produce even run-time exceptions).
+Static type checking allows us to find these errors during compilation.
 Listing \ref{oops-compilation} shows the result of attempting to compile this code.
-The compilation script completes (so that functions can refer to each other mutually recursively), then the Ace functions are typechecked.
-The typechecker raises an exception at the first error, and \verb|acec| prints it.
+The compilation script completes (so that functions can refer to each other mutually recursively), then the typed functions in the top-level environment are typechecked.
+The typechecker raises an exception statically at the first error, and \verb|acec| prints it as shown.
 \begin{codelisting}[t]
 \begin{lstlisting}
-from listing1 import A, C, log_transfer
+from listing1 import py, A, C, log_transfer
 from datetime import date
@@
               account_num: None,
-              routing_num: "0-0000-0001"} (C)
+              routing_num: "0-0000-0002"} (C)
     log_transfer((common, a))
@@
 
-print date.today().day
+print str(date.today().day)
 \end{lstlisting}
@@
 
-\section{Active Type Synthesis and Translation}\label{att}
-\todo{To be revised from here on down}The core of Ace consists of about 1500 lines of Python code implementing its primary concepts: generic functions, concrete functions, active types, active bases and active targets.
-The latter three comprise Ace's extension mechanism.
-Extensions provide semantics to, and govern the compilation of, Ace functions, rather than logic in Ace's core.
-%Indeed, the name ``Ace'' might itself be an acronym: it is an ``active compilation environment''.
-
-\subsection{Active Types}\label{atypes}
-Active types are the primary means for extending Ace with new abstractions.
-An active type, as mentioned previously, is an instance of a class implementing the \verb|ace.ActiveType| interface.
-Listing \ref{cplx} shows an example of such a class: the \verb|clx.Cplx| class used in Listing \ref{compscript}, which implements the logic of complex numbers.
-The constructor takes as a parameter any numeric type in \verb|clx| (line \ref{cplx}.2).
-
-\subsubsection{Dispatch Protocol}
+\section{Active Typechecking and Translation}\label{att}
+
+In tooling for a monolithic language, there is typically a \emph{syntax-directed} protocol governing type synthesis/analysis (collectively, \emph{typechecking}\todo{rename it above}), translation, pretty-printing and so on.
+For example, one might declare functional datatypes that capture all types and term forms, and then proceed by  exhaustive case analysis to implement these functions.
+The external visitor pattern for class-based object systems captures essentially the same protocol.
+In both cases, the problem is that all the expression forms and types are defined in one place, so one cannot extend the language modularly.
+
+In extensible languages the focus is typically on enabling the addition of new types and term forms.
+Adding new term forms requires then adding cases for all possible functions over terms (e.g.
+typing, translation,  code highlighting, style checking and so on).
+This is difficult (we will discuss some approaches in Sec.
+\ref{related}).
+Doing so while retaining the ability to reason compositionally about metatheoretic issues (e.g.
+type safety) is even more difficult.
+This is the canonical example of the widely-studied \emph{expression problem} \cite{wadler1998expression}.
+
+Ace's term forms, as we have seen, are fixed by Python's grammar, so Ace sidesteps the expression problem entirely (and thus reaps the extrinsic benefits related to tool support, described earlier).
+Instead of taking a syntax-directed view of extensibility, we take a type-directed view: users cannot add new forms but they can add new types, which can affect the semantics of existing forms according to an unambiguous protocol.
+%In Ace (but not in the theory in Sec.
+\ref{theory}), users can also repurpose existing forms flexibly within a delimited scope.
+More specifically, the key mechanisms that distinguish Ace's approach to extensibility are: \begin{itemize}
+\item a \emph{dispatch protocol} that delegates responsibility for typechecking and translation to:
+  \begin{itemize}
+  \item a type or type constructor extracted from a subterm of an expression in some way
+  \item a per-function base semantics (or simply \emph{base}) for base forms for which this is not possible  (statements, variables and unascribed literals)
+  \end{itemize}
+\item a mechanism for allowing the user to define new types, type constructors and bases from within the language (in particular, from within the type-level language) rather than fixing them ahead of time
+\end{itemize}
 \begin{codelisting}
-\lstinputlisting{listing9.py}
-\caption{\texttt{[}in \texttt{examples/clx.py]} The active type family \texttt{Ptr} implements the semantics of OpenCL pointer types.}
-\label{cplx}
-\end{codelisting}
-In a compiler for a monolithic language, there would be a \emph{syntax-directed} protocol governing typechecking and translation.
-In a compiler written in a functional language, for example, one would declare datatypes that captured all forms of types and expressions, and the typechecker would perform exhaustive case analysis over the expression forms.
-That is, all the semantics are implemented in one place.
-The visitor pattern typically used in object-oriented languages implements essentially the same protocol.
-This does not work for an extensible language because new cases and logic need to be added \emph{modularly}, in a safely composable manner.
-
-Instead of taking a syntax-directed approach, the Ace compiler's typechecking and translation phases take a \emph{type-directed approach}.
+\begin{lstlisting}
+from listing1 import py, A
+@py
+def example(a):
+  {a : A}
+  x = a.amount
+  return x
+\end{lstlisting}
+\caption{[\texttt{listing6.py}] The example detailed in Sec. \ref{att}.}
+\label{att-example}
+\end{codelisting}
+
+To see how this works, let us trace through how Ace processes the simple example in Listing \ref{att-example}.
+
+\subsection{Base Decorators}
+Function decorators in Python (line 2) are syntactic sugar for \emph{calling} the decorator (\verb|py|) with the decorated function (lines 3-5) as an argument, then replacing the binding (\verb|example|)  with the return value of this call.
+We could equivalently have omitted line 3 and inserted the top-level statement \verb|example = py(example)| on the line after the function definition.
+Note that \verb|py| is not a function but an instance of a class, \verb|examples.py.PyBase|, inheriting from \verb|ace.Base|.
+It can be used as a decorator because \verb|ace.Base| overloads the call operator as shown in Listing \ref{base}.
+The \verb|_aceify| function (not shown) \todo{show in appendix?}operates as follows:
+\begin{enumerate}
+\item The Python standard library is used to extract an abstract syntax tree (\verb|AST|) from the function.
+\item The closure of the function, together with the globals in the module it is defined in, are extracted to form its static environment.
+\item The argument annotations are processed.
+If provided in the body, as in our example, it checks that the argument names are spelled correctly and evaluates the type-level expressions (e.g.
+\verb|A| above) in the static environment to produce a mapping of argument names to types, called the argument signature.
+If Python 3's argument annotations are used, this has  already been done for us.
+%\item Whether the function is generic or not is determined.  We defer discussion of generic functions until Sec. \ref{targets}.
+\end{enumerate}
+By the end of Listing \ref{att-example}, \verb|example| is thus an instance of \verb|ace.TypedFn|.
+It could  have been constructed directly by calling the constructor on line 2 of Listing \ref{base} (this would be inconvenient, but could be useful for metaprogramming).
+
+\subsection{Typechecking}
+When the Ace compiler begins typechecking a \verb|TypedFn| (when asked to by \verb|acec|, or when it is  invoked interactively as we will show in Sec. \ref{targets}), it proceeds as follows:
+\begin{itemize}
+\item An \verb|ace.Context| is constructed with a reference to the function and the target being used  (Listing \ref{context} lines 2-4).
+A base can define a default target (Listing \ref{pybase} line 2).
+Targets are discussed further below.
+\item The base is asked to initialize the context.
+This can be seen on lines 5-7 of Listing \ref{pybase}, where the Python base adds an attribute tracking local bindings and the return type to be synthesized (initially not known).
+\item The documentation string and argument annotations, if present, are skipped (having already been processed).
+\item For each statement in the remainder of the body, the compiler calls a method of the base, \verb|check_|$F$, where $F$ is the form of the statement (taken from Python's \verb|ast| module, which performs parsing  \cite{python}).
+\end{itemize}
+........
+
 When encountering a compound term (e.g.
-\verb|e[e1]|), the compiler defers control over typechecking and translation to the active type of a designated subexpression (e.g.
-\verb|e|) determined by Ace's fixed \emph{dispatch protocol}.
-Below are examples of the choices made in Ace.
-Due to space constraints, we do not show the full protocol.
+\verb|t.amount| in Listing \ref{example}), the compiler defers control over type  and translation to the type of a designated subexpression (e.g.
+\verb|t|) according to a fixed fixed \emph{dispatch protocol}.
+Below are examples of the choices made in Ace.\todo{full in appendix?}
 \begin{itemize}
@@
 \verb|e1 + e2|) is first handed to the type assigned to the left operand.
-If it indicates a type error, the type assigned to the right operand is handed responsibility, via a different method call.
+If it indicates that it cannot handle the operation, the type assigned to the right operand is handed responsibility.
 {Note that this operates like the corresponding rule in Python's \emph{dynamic} operator overloading mechanism; see Sec.
@@
 \end{itemize}
+%
+%\todo{To be revised from here on down}The core of Ace consists of about 1500 lines of Python code implementing its primary concepts: generic functions, concrete functions, active types, active bases and active targets.
+The latter three comprise Ace's extension mechanism.
+Extensions provide semantics to, and govern the compilation of, Ace functions, rather than logic in Ace's core.
+%Indeed, the name ``Ace'' might itself be an acronym: it is an ``active compilation environment''.
+
+\subsection{Active Types}\label{atypes}
+\begin{codelisting}
+\begin{lstlisting}
+class TypedFn(object):
+  def __init__(self, base, ast, static_env, arg_sig):
+    # ...
+  # ...
+class Base(object):
+  def __call__(self, f):
+    (ast, static_env, arg_sig) = _aceify(f)
+    return TypedFn(self, ast, static_env, arg_sig)
+  # ...
+\end{lstlisting}
+\caption{A portion of the \texttt{ace} core showing how a base can be used as a decorator to construct a concrete function.}
+\label{base}
+\end{codelisting}
+\begin{codelisting}
+\begin{lstlisting}
+class PyBase(ace.Base):
+  default_target = PyTarget()
+  default_Num_ascription = int_t
+  default_Dict_ascription = dict_t
+  # ...
+  def init_ctx(self, ctx):
+    ctx.locals = { }
+    ctx.return_t = None
+
+  def check_Assign(self, ctx, s):
+    if len(s.targets) > 1:
+      # we do not consider destructuring assignment
+      raise TypeError("...not yet supported...", s)
+    x, e = s.targets[0].id, s.value # see ast module
+	if x in ctx.locals:
+	  ctx.analyze(e, ctx.locals[x])
+	else:
+	  ty = ctx.synthesize(e)
+	  ctx.locals[x] = ty
+
+  def check_Return(self, ctx, s):
+    if ctx.return_t == None:
+      ctx.return_t = ctx.synthesize(s.value)
+    else:
+      ctx.analyze(s.value, ctx.return_t)
+
+  def synthesize_Name(self, ctx, e):
+    x = e.id
+    try:
+      return ctx._locals[x]
+    elif x in ctx.fn.static_env:
+      return self.synthesize_lifted(
+        ctx.fn.static_env[x])
+    else:
+      raise ace.TypeError("...var not bound...", e)
+
+  def synthesize_FunctionDef(self, ctx):
+    if ctx.return_t == None:
+      ctx.return_t = unit
+    return arrow[ctx.fn.arg_sig, ctx.return_t]
+  # ...
+py = PyBase()
+\end{lstlisting}
+\caption{A portion of the base used in our examples thus far, defined in the \texttt{examples.py} package.}
+\label{pybase}
+\end{codelisting}
+\begin{codelisting}
+\begin{lstlisting}
+class Context(object):
+  def __init__(self, fn, target):
+    self.fn = fn
+    self.target = target
+
+  def analyze(self, e, ty):
+    if instanceof(e, ast.Num):
+      return ty.analyze_Num(self, e)
+    elif instanceof(e, ast.Str):
+      return ty.analyze_Str(self, e)
+    elif instanceof(e, ast.Dict): # TODO
+      return ty.analyze_Dict(self, e)
+    # ...
+    else:
+      syn_type = self.synthesize(e)
+      if ty != syn_ty:
+        raise TypeError("...", e)
+
+  def synthesize(self, e):
+    if instanceof(e, ast.Num):
+      self.analyze(self, e, self.fn.base.default_Num_t)
+    ...
+    elif instanceof(e, ast.Subscript):
+      t = self.synthesize(e.xxx)
+      return t.synthesize_Subscript(self, e)
+    ...
+\end{lstlisting}
+\caption{CONTXT}
+\label{context}
+\end{codelisting}
+
+Active types are the primary means for extending Ace with new abstractions.
+An active type, as mentioned previously, is an instance of a class implementing the \verb|ace.ActiveType| interface.
+Listing \ref{cplx} shows an example of such a class: the \verb|clx.Cplx| class used in Listing \ref{compscript}, which implements the logic of complex numbers.
+The constructor takes as a parameter any numeric type in \verb|clx| (line \ref{cplx}.2).
+
+\subsubsection{Dispatch Protocol}
+\begin{codelisting}
+\lstinputlisting{listing9.py}
+\caption{\texttt{[}in \texttt{examples/clx.py]} The active type family \texttt{Ptr} implements the semantics of OpenCL pointer types.}
+\label{cplx}
+\end{codelisting}
 
@@
 We then provide a different type assignment to generate the concrete function \verb|map_neg_ci32|.
-Concrete functions are instances of \verb|ace.ConcreteFn|, consisting of an abstract syntax tree annotated with types and translations along with a reference to the original generic function.
+Concrete functions are instances of \verb|ace.TypedFn|, consisting of an abstract syntax tree annotated with types and translations along with a reference to the original generic function.
 %The typing and translation process is mediated by the logic in the active base, types and target that have been provided, as we will describe in more detail below.

### 2014-03-24 ace-oopsla14/ace-oopsla14.tex
OOPSLA14 progress

@@
 \begin{codelisting}
-\lstinputlisting{listing9.py}
-\caption{\texttt{[}in \texttt{examples/clx.py]} The active type family \texttt{Ptr} implements the semantics of OpenCL pointer types.}
-\label{cplx}
-\end{codelisting}
-
+\begin{lstlisting}
+class RecordType(ace.Type):
+  def __init__(self, sig):
+    self.sig = sig
+
+  def analyze_Dict(self, ctx, e, ty):
+    ...
+
+  def translate_Dict(self, ctx, e):
+    ...
+
+  def synthesize_Attribute(self, ctx, e): 
+    ...
+
+  def translate_Attribute(self, ctx, e):
+    ...
+
+  def trans_type(self, target):
+    if target.supports(examples.fp.EagerTuple):
+      return target.EagerTuple()
+    else if isinstance(target, examples.py.Dyn):
+      return target.Dyn()
+
+record = ace.util.SlicesToOrderedDict(RecordType, 
+  TypeFormationError("..."))
+\end{lstlisting}
+\caption{Records...}
+\label{record}
+\end{codelisting}
 \subsubsection{Typechecking}
@@
 
+\appendix
+\section{Complex Numbers}
+\begin{codelisting}
+\lstinputlisting{listing9.py}
+\caption{\texttt{[}in \texttt{examples/clx.py]} The active type family \texttt{Ptr} implements the semantics of OpenCL pointer types.}
+\label{cplx}
+\end{codelisting}
+
 \end{document}

### 2014-03-24 ace-oopsla14/ace-oopsla14.tex
OOPSLA14 more progress

--- 2014-03-24 ace-oopsla14/ace-oopsla14.tex
@@
 Python serves as Ace's type-level language.
-Rather than building in a particular type system, Ace has an extensible bidirectional type system that inverts control over typing and translation to type-level functions according to a protocol that cannot lead to ambiguities and type safety issues when type systems are combined.
+Rather than building in a particular type system, Ace is organized around an extensible bidirectional type system that inverts control over typechecking and translation to type-level functions according to a protocol that cannot lead to ambiguities and type safety issues when type systems are combined.
 We give a simplified calculus that describes the type theoretic foundations of this mechanism and show that, as implemented in Ace, it is flexible enough to admit practical library-based implementations of types drawn from functional, object-oriented, parallel and domain-specific languages.
@@
 In the latter case, the ``dynamic type'' of a Python value determines a static type ``just-in-time'' when a call occurs.
-This represents a form of implicit \emph{staging} and significantly streamlines interactive workflows (common in, for example, scientific computing) that alternate between Python for orchestration and productivity-oriented tasks (e.g.
-exploratory plotting) and an external language via a foreign-function interface (FFI) for performance- and correctness-critical portions.%We will discuss in detail as our primary realistic case study.
+This represents a form of implicit \emph{staging} and significantly streamlines interactive workflows (like those in scientific computing) that alternate between Python for orchestration and exploratory tasks (e.g.
+plotting) and an external language via a foreign-function interface (FFI) for performance- and correctness-critical portions.%We will discuss in detail as our primary realistic case study.
 
@@
 \item On lines 8-11, we construct another record type.
-The field \verb|name| has type \verb|string|, defined in \verb|examples.py|, while the fields \verb|account_num| and \verb|routing_num| have more interesting types, classifying strings guaranteed statically to be in the language of a regular expression  (written, to avoid needing to escape backslashes, using Python's \emph{raw string literals}).
-Our appendix specifies the typechecking rules, based on \cite{fulton-thesis}.
+The field \verb|name| has type \verb|string|, defined in \verb|examples.py|, while the fields \verb|account_num| and \verb|routing_num| have more interesting types, classifying strings guaranteed statically to be in a regular language specified statically by a  regular expression pattern (written, to avoid needing to escape backslashes, using Python's \emph{raw string literals}).
+The appendix specifies the typechecking rules, based on \cite{fulton-thesis}.
+To our knowledge, this type system has not previously been implemented.
 We include it to emphasize that we aim to support specialized type systems,  not just general purpose constructs like records, numbers and strings.
@@
 %Prototypic inheritance  is a feature of some dynamically-typed languages (most notably, Self \cite{Ungar:Smith:oopsla:1987} and Javascript). 
-On line 12, we consider a simple \emph{prototypic object type} \cite{Lie86}.
+On line 12, we construct a simple \emph{prototypic object type} \cite{Lie86}.
 The type \verb|Transfer| classifies terms consisting of a \todo{what is the right term for this?}\emph{fore} of type \verb|A| and a \emph{prototype} of type \verb|C|.
@@
 
-Line 17 borrows Python's syntax for dictionary literals to approximate  conventional notation for type annotations.\footnote{In the rare situation where there are no type annotations and the first statement is a dictionary literal, users can use Python's \texttt{pass} keyword.} % (and following conventions for documentation strings, which also describe ``types'' at the end).
+Line 17 borrows Python's syntax for dictionary literals to approximate  conventional notation for type annotations.
+%\footnote{In the rare situation where there are no type annotations and the first statement is a dictionary literal, users can use Python's \texttt{pass} keyword.} % (and following conventions for documentation strings, which also describe ``types'' at the end).
 % The ``keyword'' \verb|type| was chosen because it is the name of a built-in function in Python (and thus would not be used in the way shown for any other purpose).
- {In version 3.0 of Python, syntax for annotating arguments with arbitrary values directly was introduced \cite{PEP}:
+ {In version 3.0 of Python, syntax for annotating arguments with arbitrary values directly was introduced \cite{pep3107}:
 % \vspace{-5px}
@@
 from datetime import date
-
 @py
@@
     a.amount += 1000 
-
 print str(date.today().day)
@@
 Ace's term forms, as we have seen, are fixed by Python's grammar, so Ace sidesteps the expression problem entirely (and thus reaps the extrinsic benefits related to tool support, described earlier).
-Instead of taking a syntax-directed view of extensibility, we take a type-directed view: users cannot add new forms but they can add new types, which can affect the semantics of existing forms according to an unambiguous protocol.
+Instead of taking a syntax-directed view of extensibility, we take a type-directed view: users cannot add new forms but they can add new types, which can affect the semantics of existing forms by an unambiguous protocol.
 %In Ace (but not in the theory in Sec.
@@
   x = a.amount
-  return x
+  x = x + x
+  return {amount: x} (A)
 \end{lstlisting}
@@
 
-\subsection{Typechecking}
-When the Ace compiler begins typechecking a \verb|TypedFn| (when asked to by \verb|acec|, or when it is  invoked interactively as we will show in Sec. \ref{targets}), it proceeds as follows:
-\begin{itemize}
-\item An \verb|ace.Context| is constructed with a reference to the function and the target being used  (Listing \ref{context} lines 2-4).
-A base can define a default target (Listing \ref{pybase} line 2).
-Targets are discussed further below.
-\item The base is asked to initialize the context.
-This can be seen on lines 5-7 of Listing \ref{pybase}, where the Python base adds an attribute tracking local bindings and the return type to be synthesized (initially not known).
-\item The documentation string and argument annotations, if present, are skipped (having already been processed).
-\item For each statement in the remainder of the body, the compiler calls a method of the base, \verb|check_|$F$, where $F$ is the form of the statement (taken from Python's \verb|ast| module, which performs parsing  \cite{python}).
-\end{itemize}
-........
-
-When encountering a compound term (e.g.
-\verb|t.amount| in Listing \ref{example}), the compiler defers control over type  and translation to the type of a designated subexpression (e.g.
-\verb|t|) according to a fixed fixed \emph{dispatch protocol}.
-Below are examples of the choices made in Ace.\todo{full in appendix?}
-\begin{itemize}
-\item Responsibility over {\bf attribute access} (\texttt{e.attr}), {\bf subscripting} (\texttt{e[e1]}) and \textbf{calls} (\verb|e(e1, ..., en)|) and {\bf unary operations} (e.g.
-\verb|-e|) is handed to the type recursively assigned to \texttt{e}.
-\item Responsibility over {\bf binary operations} (e.g.
-\verb|e1 + e2|) is first handed to the type assigned to the left operand.
-If it indicates that it cannot handle the operation, the type assigned to the right operand is handed responsibility.
-{Note that this operates like the corresponding rule in Python's \emph{dynamic} operator overloading mechanism; see Sec.
-\ref{related} for a discussion.}
-\item Responsibility over \textbf{constructor calls} (\verb|[t](e1, ..., en)|), where \verb|t| is a \emph{compile-time Python expression} evaluating to an active type, is handed to that type.
-If \verb|t| evaluates to a \emph{family} of types, like \verb|clx.Cplx|, the active type is first generated via a class method, as discussed below.
-%\item Responsibility over {\bf simple assignment statements}, is handed to the type of the variable on the left (which, as we will see below, is determined by the active base).
-If this type does not provide special assignment semantics, the base must handle it.%Destructuring assignment is also supported by a somewhat more complex protocol.\todo{clarify}
-\end{itemize}
-%
-%\todo{To be revised from here on down}The core of Ace consists of about 1500 lines of Python code implementing its primary concepts: generic functions, concrete functions, active types, active bases and active targets.
-The latter three comprise Ace's extension mechanism.
-Extensions provide semantics to, and govern the compilation of, Ace functions, rather than logic in Ace's core.
-%Indeed, the name ``Ace'' might itself be an acronym: it is an ``active compilation environment''.
-
-\subsection{Active Types}\label{atypes}
+\subsection{Typechecking Typed Functions and Statements}
 \begin{codelisting}
@@
 \end{lstlisting}
-\caption{A portion of the \texttt{ace} core showing how a base can be used as a decorator to construct a concrete function.}
+\caption{A portion of the \texttt{ace} core showing how a base can be used as a decorator to construct a typed function.}
 \label{base}
 \end{codelisting}
+When the Ace compiler begins typechecking a \verb|TypedFn| (when asked to by \verb|acec|, or when it is  invoked interactively as we will show in Sec. \ref{targets}), it proceeds as follows:
+\begin{enumerate}
+\item An \verb|ace.Context| is constructed with a reference to the function (Listing \ref{context} lines 2-3).
+\item The base is asked to initialize the context.
+This can be seen on lines 5-7 of Listing \ref{pybase}, where the Python base adds an attribute tracking local bindings (initially only the arguments) and the return type to be synthesized (initially not known, indicated using \verb|None|).
+\item For each statement in the remainder of the body, the compiler calls a method of the base, \verb|check_|$F$, where $F$ is the form of the statement (taken directly from Python's \verb|ast| package \cite{python}).
+These methods are responsible for checking that statement is well-typed (statements do not themselves have a type) and implementing any necessary effects on the context.
+\item The base synthesizes a type for the function as a whole via the \verb|syn_ty_FunctionDef_outer| method.
+\end{enumerate}
+
+The assignment statement on lines 5 and 6 of Listing \ref{att-example} triggers the method \verb|check_Assign| \todo{if time, add destructuring assignment to appendix}shown on lines 6-14 of Listing \ref{pybase} and the return statement on line 7 triggers the method \verb|check_Return| on lines 18-22.
+Both work similarly: if this is the first assignment to a particular identifier, or the first return statement seen, then the value must synthesize a type in the current context.
+Otherwise, the value is analyzed against the previously determined type.
+At the end, an \verb|arrow| type is synthesized (lines 26-29) based on the argument signature and the synthesized return type.
+
+\subsection{Bidirectionally Typechecking Expressions}
 \begin{codelisting}
@@
 class PyBase(ace.Base):
-  default_target = PyTarget()
-  default_Num_ascription = int_t
-  default_Dict_ascription = dict_t
-  # ...
   def init_ctx(self, ctx):
-    ctx.locals = { }
+    ctx.locals = dict(ctx.fn.arg_sig)
     ctx.return_t = None
@@
     if len(s.targets) > 1:
-      # we do not consider destructuring assignment
-      raise TypeError("...not yet supported...", s)
+      raise ace.TypeError("...not supported...", s)
     x, e = s.targets[0].id, s.value # see ast module
 	if x in ctx.locals:
-	  ctx.analyze(e, ctx.locals[x])
+	  ctx.ana_ty(e, ctx.locals[x])
 	else:
-	  ty = ctx.synthesize(e)
+	  ty = ctx.syn_ty(e)
 	  ctx.locals[x] = ty
+
+  def trans_Assign(self, ctx, target, s):
+    return target.direct_translation(ctx, s)
 
@@
     if ctx.return_t == None:
-      ctx.return_t = ctx.synthesize(s.value)
+      ctx.return_t = ctx.syn_ty(s.value)
     else:
-      ctx.analyze(s.value, ctx.return_t)
-
-  def synthesize_Name(self, ctx, e):
-    x = e.id
-    try:
-      return ctx._locals[x]
-    elif x in ctx.fn.static_env:
-      return self.synthesize_lifted(
-        ctx.fn.static_env[x])
-    else:
-      raise ace.TypeError("...var not bound...", e)
-
-  def synthesize_FunctionDef(self, ctx):
+      ctx.ana_ty(s.value, ctx.return_t)
+
+  def trans_Return(self, ctx, target, s):
+    return target.direct_translation(ctx, s)
+
+  def syn_ty_FunctionDef_outer(self, ctx, f):
     if ctx.return_t == None:
@@
     return arrow[ctx.fn.arg_sig, ctx.return_t]
+
+  def trans_FunctionDef_outer(self, ctx, target, f):
+    if f.name == "__toplevel__":
+      return target.Suite(ctx.trans(f.body))
+    else:
+      return target.direct_translation(f, ctx)
+
+  def syn_ty_Name(self, ctx, e):
+    x = e.id
+    if x in ctx.locals:
+      return ctx.locals[x]
+    elif x in ctx.fn.static_env: # (Sec. 5)
+      return self.syn_ty_lifted(x)
+    else:
+      raise ace.TypeError("...var not bound...", e)
+
+  def trans_Name(self, ctx, target, e):
+    if e.id in ctx.locals:
+      return target.direct_translation(ctx, e)
+    else: # (Sec. 5)
+      return self.trans_lifted(ctx, target, e)
+  # ...
+  default_target = PyTarget()   
+  default_Dict_ascription = dyn
   # ...
@@
 \end{codelisting}
+
+The novel protocol for typechecking expressions that we will now describe (and formalize in Sec.
+\ref{theory}) is what enables us to extend the language with new semantics modularly while sidestepping the expression problem.
+Let us look more closely at how lines \ref{att-example}.5-\ref{att-example}.7\footnote{In this section, we will need to refer to code in Listings \ref{att-example}, \ref{pybase}, \ref{context} and \ref{record}, so we adopt this syntactic convention to refer to line numbers for concision.} are typechecked.
+
+As stated above, line \ref{att-example}.5 is an assignment statement, so the method \verb|check_Assign| on lines \ref{pybase}.6-\ref{pybase}.14 is called.
+Because the locals dictionary in \verb|ctx| does not yet know what type \verb|x| should be, the base asks to synthesize a type for the expression being bound, \verb|a.amount|, by calling the \verb|ctx.syn_ty| method.
+
+\paragraph{Active Type Synthesis} This method is defined on lines \ref{context}.5-\ref{context}.26.
+
+\subsection{Translation}
+The decorator, documentation string and argument annotations, if present, are stripped away (having already been processed).
+
+Once typechecking a method is complete, the compiler enters the translation phase, where terms in the target language are generated from Ace terms.
+Terms in the target language are generated by calling methods of the \emph{active target} governing the function being compiled.
+The translation phase operates similarly to typechecking, using the dispatch protocol to invoke methods named \verb|trans_|$X$.
+These methods have access to the context and node just as during typechecking, as well as the active target (named \verb|target| here).
+
+%When encountering a compound term (e.g.
+\verb|t.amount| in Listing \ref{example}), the compiler defers control over type  and translation to the type of a designated subexpression (e.g.
+\verb|t|) according to a fixed fixed \emph{dispatch protocol}.
+Below are examples of the choices made in Ace.\todo{full in appendix?}
+%\begin{itemize}
+%\item Responsibility over {\bf attribute access} (\texttt{e.attr}), {\bf subscripting} (\texttt{e[e1]}) and \textbf{calls} (\verb|e(e1, ..., en)|) and {\bf unary operations} (e.g.
+\verb|-e|) is handed to the type recursively assigned to \texttt{e}.
+%\item Responsibility over {\bf binary operations} (e.g.
+\verb|e1 + e2|) is first handed to the type assigned to the left operand.
+If it indicates that it cannot handle the operation, the type assigned to the right operand is handed responsibility.
+{Note that this operates like the corresponding rule in Python's \emph{dynamic} operator overloading mechanism; see Sec.
+\ref{related} for a discussion.}
+%\item Responsibility over \textbf{constructor calls} (\verb|[t](e1, ..., en)|), where \verb|t| is a \emph{compile-time Python expression} evaluating to an active type, is handed to that type.
+If \verb|t| evaluates to a \emph{family} of types, like \verb|clx.Cplx|, the active type is first generated via a class method, as discussed below.
+%%\item Responsibility over {\bf simple assignment statements}, is handed to the type of the variable on the left (which, as we will see below, is determined by the active base).
+If this type does not provide special assignment semantics, the base must handle it.%Destructuring assignment is also supported by a somewhat more complex protocol.\todo{clarify}
+%\end{itemize}
+%%
+%%\todo{To be revised from here on down}The core of Ace consists of about 1500 lines of Python code implementing its primary concepts: generic functions, concrete functions, active types, active bases and active targets.
+The latter three comprise Ace's extension mechanism.
+Extensions provide semantics to, and govern the compilation of, Ace functions, rather than logic in Ace's core.
+%Indeed, the name ``Ace'' might itself be an acronym: it is an ``active compilation environment''.
+
+\subsection{Active Types}\label{atypes}
+
 \begin{codelisting}
@@
 class Context(object):
-  def __init__(self, fn, target):
+  def __init__(self, fn):
     self.fn = fn
-    self.target = target
-
-  def analyze(self, e, ty):
-    if instanceof(e, ast.Num):
-      return ty.analyze_Num(self, e)
-    elif instanceof(e, ast.Str):
-      return ty.analyze_Str(self, e)
-    elif instanceof(e, ast.Dict): # TODO
-      return ty.analyze_Dict(self, e)
-    # ...
+
+  def syn_ty(self, e):
+    if instanceof(e, ast.Name):
+      owner = self.fn.base
+      ty = owner.syn_ty_Name(self, e)
+    elif instanceof(e, ast.Attribute):
+      owner = self.syn_ty(e.value)
+      ty = owner.syn_ty_Attribute(self, e)
+    # ... other compound forms similar (cf appendix)
+    elif isinstance(e, ast.Dict):
+      return self.ana_ty(self, e, 
+        self.fn.base.default_Dict_ty)
+    # ... other literal forms similar
+    elif is_ascribed_Num(e):
+      owner = get_ascription(e)
+      if is_tycon(owner):
+        ty = owner.syn_ty_Num(self, e)
       else:
-      syn_type = self.synthesize(e)
+        return owner.ana_ty_Num(self, e)
+    # ... other ascribed literal forms similar
+    e.owner = owner
+    e.ty = ty
+    return ty
+
+  def ana_ty(self, e, ty):
+    if instanceof(e, ast.Dict):
+      ty.ana_ty_Dict(self, e)
+    # ... other literal forms similar
+    else:
+      syn_ty = self.syn_ty(e)
       if ty != syn_ty:
-        raise TypeError("...", e)
-
-  def synthesize(self, e):
-    if instanceof(e, ast.Num):
-      self.analyze(self, e, self.fn.base.default_Num_t)
-    ...
-    elif instanceof(e, ast.Subscript):
-      t = self.synthesize(e.xxx)
-      return t.synthesize_Subscript(self, e)
-    ...
+        raise TypeError("...syn/ana mismatch...", e)
+      return
+    e.owner = e.ty = ty
+
+  def trans(self, target, e):
+    if instanceof(e, ast.Name): 
+      trans = e.owner.trans_Name(self, target, e)
+    elif instanceof(e, ast.Attribute):
+      trans = e.owner.trans_Attribute(self, target, e)
+	# ... other forms similar
+	e.trans = trans
+	return trans
 \end{lstlisting}
-\caption{CONTXT}
+\caption{The context dispatches active typechecking and translation of expressions to the base or an appropriate type or type constructor.}
 \label{context}
@@
 
-Active types are the primary means for extending Ace with new abstractions.
-An active type, as mentioned previously, is an instance of a class implementing the \verb|ace.ActiveType| interface.
-Listing \ref{cplx} shows an example of such a class: the \verb|clx.Cplx| class used in Listing \ref{compscript}, which implements the logic of complex numbers.
+%Active types are the primary means for extending Ace with new abstractions.
+An active type, as mentioned previously, is an instance of a class implementing the \verb|ace.Type| interface.
+Listing \ref{record} shows an example of such a class: the \verb|clx.Cplx| class used in Listing \ref{compscript}, which implements the logic of complex numbers.
 The constructor takes as a parameter any numeric type in \verb|clx| (line \ref{cplx}.2).
@@
 \begin{codelisting}
-\lstinputlisting{listing9.py}
-\caption{\texttt{[}in \texttt{examples/clx.py]} The active type family \texttt{Ptr} implements the semantics of OpenCL pointer types.}
-\label{cplx}
-\end{codelisting}
-
+\begin{lstlisting}
+@slices_to_sig
+class record(ace.Type):
+  def __init__(self, sig):
+    _check_sig(sig)
+    self.sig = sig
+
+  def ana_ty_Dict(self, ctx, e):
+    for f, v in zip(e.keys, e.values):
+      if f.id in self.sig:
+        ctx.ana_ty(v, self.sig[f.id])
+      else:
+        raise ace.TypeError("...bad field...", f)
+    if len(self.sig) != len(e.keys):
+      raise ace.TypeError("...missing field...", e)
+
+  def trans_Dict(self, ctx, target, e):
+    if len(self.sig) == 1: # assuming non-empty
+      return ctx.trans(e.values[0])
+    else:
+      value_dict = dict(zip(e.keys, e.values))
+	  return target.Tuple(
+	    ctx.trans(target, value_dict[field])
+	    for field, ty in self.sig)
+
+  def syn_ty_Attribute(self, ctx, e): 
+    if e.attr in self.sig: 
+      return self.sig[e.attr]
+    else:
+      raise ace.TypeError("...field not found...", e)
+
+  def trans_Attribute(self, ctx, target, e): 
+    idx = idx_of(self.sig, e.attr)
+    return target.Subscript(
+      ctx.trans(target, e.value), target.Num(idx))
+\end{lstlisting}
+%
+%  def trans_type(self, target):
+%    if target.supports(examples.fp.EagerNAryProdLR):
+%      return target.EagerNAryProdLR()
+%    else if isinstance(target, examples.py.Dyn):
+%      return target.Dyn()
+\caption{The \texttt{record} type constructor.}
+\label{record}
+\end{codelisting}
 \subsubsection{Typechecking}
 When typechecking a compound expression or statement, the Ace compiler temporarily hands control to the object selected by the dispatch protocol by calling the method \verb|type_|$X$, where $X$ is  the name of the syntactic form, taken from the Python grammar \cite{python} (appended with a suffix in some cases).
-
-For example, if \verb|c| is a complex number, then \verb|c.ni| and \verb|c.i| are its non-imaginary and imaginary components, respectively.
-These expressions are of the form \verb|Attribute|, so the typechecker calls \verb|type_Attribute| (line \ref{cplx}.7).
-This method receives the compilation context, \verb|context|, and the abstract syntax tree of the expression, \verb|node| and must return a type assignment for the node, or raise an \verb|ace.TypeError| if there is an error.
-In this case, a type assignment is possible if the attribute name is either \verb|"ni"| or \verb|"i"|, and an error is raised otherwise (lines \ref{cplx}.8-\ref{cplx}.10).
-We note that error messages are an important and sometimes overlooked facet of {ease-of-use} \cite{marceau2011measuring}.
-A common frustration with using general-purpose abstraction mechanisms to encode an abstraction is that they can produce  verbose and cryptic error messages that reflect the implementation details instead of the semantics.
-Ace supports custom error messages.%Active types permit direct encoding of semantics and customization of error messages.
-
-Complex numbers also support binary arithmetic operations partnered with both other complex numbers and with non-complex numbers, treating them as if their imaginary component is zero.
-The typechecking rules for this logic is implemented on lines \ref{cplx}.17-\ref{cplx}.29.
-Arithmetic operations are usually symmetric, so the dispatch protocol checks the types of both subexpressions for support.
-To ensure that the semantics remain deterministic in the case that both types support the binary operation, Ace asks the left first (via \verb|type_BinOp_left|), asking the right (via \verb|type_BinOp_right|) only if the left indicates an error.
-In either position, our implementation begins by recursively assigning a type to the other operand in the current context via the \verb|context.type| method (line \ref{cplx}.24).
-If supported, it applies the C99 rules for arithmetic operations to determine the resulting type (via \verb|c99_binop_t|, not shown).
-
-Finally, a complex number can be constructed inside an Ace function using Ace's special constructor form: \verb|[clx.Cplx](3,4)| represents $3+4i$, for example.
-The term within the braces is evaluated at \emph{compile-time}.
-Because \verb|clx.Cplx| evaluates not to an active type, but to a class, this form is assigned a type by handing control to the class object via the \emph{class method} \verb|type_New|.
-It operates as expected, extracting the types of the two arguments to construct an appropriate complex number type (lines \ref{cplx}.50-\ref{cplx}.57), raising a type error if the arguments cannot be promoted to a common type according to the rules of C99 or if two arguments were not provided.% (an exercise for the reader: modify this method to also allow a single argument for when the imaginary part is 0).
 
@@
 \subsubsection{Translation}
-Once typechecking a method is complete, the compiler enters the translation phase, where terms in the target language are generated from Ace terms.
-Terms in the target language are generated by calling methods of the \emph{active target} governing the function being compiled.
-The translation phase operates similarly to typechecking, using the dispatch protocol to invoke methods named \verb|trans_|$X$.
-These methods have access to the context and node just as during typechecking, as well as the active target (named \verb|target| here).
-
-As seen in Listing \ref{mapout}, we are implementing complex numbers internally using OpenCL vector types, like \verb|int2|.
-Let us look first at \verb|trans_New| on lines \ref{cplx}.54-\ref{cplx}.60, where new complex numbers are translated to vector literals by invoking \verb|target.VecLit|.
-This will ultimately  generate the necessary OpenCL code, as a string, to complete compilation (these strings are not directly manipulated by extensions, however, to avoid problems with, e.g.
-precedence).
-For it to be possible to reason compositionally about the correctness of compilation, all complex numbers must translate to terms in the target language that have a consistent target type.
-The \verb|trans_type| method of the \verb|ace.ActiveType| associates a type in the target language, here a vector type like \verb|int2|, with the active type.
-Ace supports a mode where this \emph{representational consistency} is dynamically checked during compilation (requiring that the active target know how to assign types to terms in the target language, which can be done  for our OpenCL target as of this writing).
-
-The translation methods for attributes (line \ref{cplx}.12) and binary operations  (line \ref{cplx}.31) proceed in a straightforward manner.
-The context provides a method, \verb|trans|, for recursively determining the translation of subexpressions as needed.
-Of note is that the translation methods can assume that typechecking succeeded.
-For example, the implementation of \verb|trans_Attribute| assumes that if \verb|node.attr| is not \verb|'ni'| then it must have been \verb|'i'| on line \ref{cplx}.14, consistent with the implementation of \verb|type_Attribute| above it.
-Typechecking and translation are separated into two methods to emphasize that typechecking is not target-dependent, and to allow for more advanced uses, like type refinements and hypothetical typing judgements, that we do not describe here.
-
-\subsection{Active Bases}\label{abases}
-Each generic function is associated with an active base, which is an object implementing the \verb|ace.ActiveBase| interface.
-The active base specifies the \emph{base semantics} of that function.
-It controls the semantics of statements and expressions that do not have a clear ``primary'' subexpression for the dispatch protocol to defer to.
-A base is handed control over typechecking of statements and expressions in the same way as active types: via \verb|type_|$X$ and \verb|trans_|$X$ methods.
-Each function can have a different base.
-
-Literals are the most prominent form given their semantics by an active base.
-Our example active base, \verb|clx.base|, assigns integer literals the type \verb|clx.int32| while floating point literals have type \verb|clx.double|, consistent with the semantics of OpenCL and C99.
-The \verb|clx.base| object is an instance of \verb|clx.Base| that is pre-constructed for convenience.
-Alternative bases can be generated via different constructor arguments.
-For example, \verb|clx.Base(flit_t=clx.float)| is a base semantics where floating point literals have type \verb|clx.float|.
-This is useful because some OpenCL devices do not support double-precision numbers, or impose a significant performance penalty for their use.
-Indeed, to even use the \verb|double| type in OpenCL, a flag must be toggled by a \verb|#pragma| (The OpenCL target we have implemented inserts this automatically when needed, along with some other flags, and annotations like \verb|kernel|).
-Similarly, for some applications, avoiding accidental overflows is more important than performance.
-Infinite-precision integers can be implemented as an active type (not shown) and the base can be asked to use it for numeric literals.
-In all cases, this occurs by inserting a constructor call form, as described above.% the constructor call form, described above, around integers to support this.
-
-The base also initializes the context and governs the semantics of variables and assignment.
-This can also permit it to allow for the use of ``built-in'' operators that were not explicitly imported.
-The semantics of \verb|get_global_id| and \verb|printf| are provided by \verb|clx.base|.
-In fact, the base provides extended versions of them (the former permits multi-dimensional global IDs, the latter supports typechecking format strings).A base which does not provide them as built-ins (requiring that they be imported explicitly) can be generated by passing the \verb|primitives=False| flag.
-
-\subsection{Active Targets and Cross-Compilation}\label{atargets}
-An active target, which we describe being used above, is an instance of \verb|ace.ActiveTarget|.
-It has two primary roles in Ace: 1) providing an API for code generation to a target language, providing term and type constructors for use during translation (and, optionally, a type system implementation to support representational consistency checks); 2) implementing the interoperability layer to support calling of  generic or concrete function directly from Python, as described in Sec.
-\ref{implicit}.
-
-An active type or base can support multiple active targets by simply examining \verb|target| during translation (e.g.
-by performing capability checks using Python's \verb|hasattr| function).
-Alternatively, the active target can simulate the interface of a different target but generate code differently.
-For example, although our implementation of OpenCL's type system as a collection of active types and an active base relies on a target that supports OpenCL terms and types, our \verb|CUDA| and \verb|C99| targets provide partial support for the same term and type forms, in order to support cross-compilation.
-We will show the latter being used in Sec.
-\ref{examples}.
 
@@
 
-\section{Expressiveness}\label{examples}
-Thus far, we have focused mainly on the OpenCL target and shown examples of fairly low-level active types: those that implement OpenCL's primitives (e.g.
-\verb|clx.Ptr|), extend them in simple but convenient ways (e.g.
-\verb|clx.Cplx|) and those that make interactive execution across language boundaries safe and convenient (e.g.
-\verb|clx.NPArray|).
-Ace was first conceived to answer the question: \emph{can we build a statically-typed language with the semantics of a C but the syntax and ease-of-use of a Python?} We submit that the work described above has answered this question in the affirmative.
-
-But Ace has proven useful for more than low-level tasks like programming a GPU with OpenCL.
-We now describe several interesting extensions that implement the semantics of primitives drawn from a range of different language paradigms, to justify our claim that these mechanisms are highly expressive.
-
-\subsection{Growing a Statically-Typed Python Inside an Ace}
-Ace comes with a target, base and type implementing Python itself: \verb|ace.python.python|, \verb|ace.python.base| and \verb|ace.python.dyn|.
-These can be supplemented by additional active types and used as the foundation for writing actively-typed Python functions.
-These functions can either be compiled ahead-of-time to an untyped Python file for execution, or be immediately executed with just-in-time compilation, just like the OpenCL examples we have shown.
-Many of the examples in the next section support this target in addition to the OpenCL target we have focused on thus far.
-
-\subsection{Recursive Labeled Sums}
-\begin{codelisting}
-\lstinputlisting{datatypes_t.py}
-\caption{\texttt{[datatypes\_t.py]} An example using statically-typed functional datatypes.}
-\label{datatypest}
-\end{codelisting}
-\begin{codelisting}
-\lstinputlisting{datatypes.py}
-\caption{\texttt{[datatypes.py]} The dynamically-typed Python code generated by running \texttt{acec datatypes\_t.py}.}
-\label{datatypes}
-\end{codelisting}
-Listing \ref{datatypest} shows an example of the use of statically-typed functional datatypes (a.k.a.
-recursive labeled sum types) together with the Python implementation just described.
-It shows two syntactic conveniences that were not mentioned previously: 1) if no target is provided to \verb|ace.fn|, the base can provide a default target (here, \verb|ace.python.python|); 2) a concrete function can be generated immediately by providing a type assignment after \verb|ace.fn| using braces.
-Listing \ref{datatypest} generates Listing \ref{datatypes}.
-
-Lines \ref{datatypest}.3-\ref{datatypest}.11 define a function that generates a recursive algebraic datatype representing a tree given a name and another Ace type for the data at the leaves.
-This type implemented by the active type family \verb|fp.Datatype|.
-A name for the datatype and the case names and types are provided programmatically on lines \ref{datatypest}.3-\ref{datatypest}.8.
-To support recursive datatypes, the case names are enclosed within a lambda term that will be passed a reference to the datatype itself.
-These lines also show two more active type families: units (the type containing just one value), and immutable pairs.
-Line \ref{datatypest}.13 calls this function with the Ace type for dynamic Python values, \verb|py.dyn|, to generate a type, aliased \verb|DT|.
-This type is implemented using class inheritance when the target is Python, as seen on lines \ref{datatypes}.1-\ref{datatypes}.13.
-For C-like targets, a \verb|union| type can be used (not shown).
-
-The generic function \verb|depth_gt_2| demonstrates two features.
-First, the base has been setup to treat the final expression in a top-level branch as the return value of the function, consistent with typical functional languages.
-Second, the \verb|case| ``method'' on line \ref{datatypest}.17 creatively reuses the syntax for dictionary literals to express a nested pattern matching construct.
-The patterns to the left of the colons are not treated as expressions (that is, a type and translation is never recursively assigned to them).
-Instead, the active type implements the standard algorithm for ensuring that the cases cover all possibilities and are not redundant \cite{pfpl}.
-If the final case were omitted, for example, the algorithm would statically indicate an error, just as in statically-typed functional languages like ML.
-
-\subsection{Nested Data Parallelism}
-Functional constructs have shown promise as the basis for a nested data-parallel parallel programming model.
-Many powerful parallel algorithms can be specified by composing primitives like \verb|map| and \verb|reduce| on persistent data structures like lists, trees and records.
-Data dependencies are directly encoded in these specifications, and automatic code generation techniques have shown promise in using this information to automatically execute these specifications on concurrent hardware and networks.
-The Copperhead programming language, for example, is based in Python as well and allows users to express functional programs over lists using common functional primitives, compiling them to CUDA code \cite{catanzaro2011copperhead}.
-By combining active types and the metaprogramming facilities described in Sec.
-2 to implement optimizations, like fusion, on the typed syntax trees available from concrete functions, these same techniques can be implemented as an Ace library as well.
-
-\subsection{Product Types}
-Product types (types that group heterogeneous values together) like structs, unions, tuples, records and objects can all be implemented as Ace type families  parameterized by a dictionary mapping field names (indices, in the case of tuples) to types.
-Each family differs slightly in the semantics of the operations available.
-Structs support mutation, for example, while records do not.
-Object systems typically introduce additional logic for calling methods, binding special variables like \verb|this|, and accessing information through an inheritance hierarchy.
-Listing \ref{ooclxfppy} shows an example of each of these.
-The prototypic object system delegates to a backing record if a field is not available in the foreground.
-This example also demonstrates cross-compilation to C99, supported for a subset of operations in \verb|clx|.
-All three abstractions are implemented as simple \verb|struct|s, as can be seen in Listing \ref{ooclxfpc}.
-This example further demonstrates the use of argument annotations to generate concrete functions (line \ref{ooclxfppy}.9), which is a feature only available when using Python 3.3+.
-
-\subsection{Distributed Programming}
-Many distributed programming abstractions can be understood as implementing object models with complex forms of dynamic dispatch.
-For example, in Charm++, dynamic dispatch involves message passing over a dynamically load-balanced network \cite{kale2009charm++}.
-While Charm++ is a sophisticated system, and we do not anticipate that implementing it as an Ace extension would be easy, it is possible to do so by targeting a system like Adaptive MPI, which exposes the Charm++ runtime system \cite{kale2009charm++}.
-%Erlang and other message-passing and actor-based systems can also be considered in this manner -- processes can be thought of as objects, channels as methods and messages as calls.
-
-%\subparagraph{Annotation and Extension Inference}
-%In addition to type annotations, OpenCL normally asks for additional annotations in a number of other situations.
-Users can annotate functions that meet certain requirements  with the \verb|__kernel| attribute, indicating that they are callable from the host.
-The \verb|OpenCL| backend can check these and add this annotation automatically.
-Several types (notably, \verb|double|) and specialized primitive functions require that an OpenCL extension be enabled via a \verb|#pragma|.
-The OpenCL backend automatically detects many of these cases as well, adding the appropriate \verb|#pragma| declaration.
-An example of this can be seen in Listing \ref{mapout}, where the use of the \verb|double| type triggers the insertion of the \verb|cl_khr_f64| extension.
-%
-
-A number of recent languages designed for distributed programming on clusters use a partitioned global address space model, e.g.
-Chapel \cite{chapel}.
-These languages provide first-class support for accessing data transparently across a massively parallel cluster.
-Their type systems track information about \emph{data locality} so that the compiler can emit more efficient code.
-The extension mechanism can also track this information in a manner analogous to how address space information is tracked in the OpenCL example above, and target an existing portable run-time such as GASNet \cite{bonachea2002gasnet}.
-
-\subsubsection{Units of Measure}
-A number of domain-specific type systems can be implemented within Ace as well.
-For example, prior work has considered tracking units of measure (e.g.
-grams) statically to validate scientific code \cite{conf/cefp/Kennedy09}.
-This cannot easily be implemented using many existing abstraction mechanisms because this information should only be maintained statically to avoid excessive run-time overhead associated with tagging/boxing, and the typechecking logic is reasonably complex.
-The Ace extension mechanism allows this information to be tracked in the type system, but not included during translation, by a mechanism similar to how address space information is tracked with pointers.
-Because Ace extensions can use Python, the needed logic can be implemented directly.
-
 \section{Related Work}\label{related}
-\begin{codelisting}[t]
-\lstinputlisting{ooclxfp.py}
-\caption{\texttt{[ooclxfp.py]} An example combining structs and immutable records using a prototype-based object system, cross-compiled to C99. Uses Python 3 argument annotations.}
-\label{ooclxfppy}
-\end{codelisting}
-\begin{codelisting}[t]
-\lstinputlisting[style=OpenCL]{ooclxfp.c}
-\caption{\texttt{[ooclxfp.c]} The C99 code generated by running \texttt{acec ooclxfp.py}.}
-\label{ooclxfpc}
-\end{codelisting}
 
@@
 
+\appendix
+\section{Complex Numbers}
+For example, if \verb|c| is a complex number, then \verb|c.ni| and \verb|c.i| are its non-imaginary and imaginary components, respectively.
+These expressions are of the form \verb|Attribute|, so the typechecker calls \verb|type_Attribute| (line \ref{cplx}.7).
+This method receives the compilation context, \verb|context|, and the abstract syntax tree of the expression, \verb|node| and must return a type assignment for the node, or raise an \verb|ace.TypeError| if there is an error.
+In this case, a type assignment is possible if the attribute name is either \verb|"ni"| or \verb|"i"|, and an error is raised otherwise (lines \ref{cplx}.8-\ref{cplx}.10).
+We note that error messages are an important and sometimes overlooked facet of {ease-of-use} \cite{marceau2011measuring}.
+A common frustration with using general-purpose abstraction mechanisms to encode an abstraction is that they can produce  verbose and cryptic error messages that reflect the implementation details instead of the semantics.
+Ace supports custom error messages.%Active types permit direct encoding of semantics and customization of error messages.
+
+Complex numbers also support binary arithmetic operations partnered with both other complex numbers and with non-complex numbers, treating them as if their imaginary component is zero.
+The typechecking rules for this logic is implemented on lines \ref{cplx}.17-\ref{cplx}.29.
+Arithmetic operations are usually symmetric, so the dispatch protocol checks the types of both subexpressions for support.
+To ensure that the semantics remain deterministic in the case that both types support the binary operation, Ace asks the left first (via \verb|type_BinOp_left|), asking the right (via \verb|type_BinOp_right|) only if the left indicates an error.
+In either position, our implementation begins by recursively assigning a type to the other operand in the current context via the \verb|context.type| method (line \ref{cplx}.24).
+If supported, it applies the C99 rules for arithmetic operations to determine the resulting type (via \verb|c99_binop_t|, not shown).
+
+Finally, a complex number can be constructed inside an Ace function using Ace's special constructor form: \verb|[clx.Cplx](3,4)| represents $3+4i$, for example.
+The term within the braces is evaluated at \emph{compile-time}.
+Because \verb|clx.Cplx| evaluates not to an active type, but to a class, this form is assigned a type by handing control to the class object via the \emph{class method} \verb|type_New|.
+It operates as expected, extracting the types of the two arguments to construct an appropriate complex number type (lines \ref{cplx}.50-\ref{cplx}.57), raising a type error if the arguments cannot be promoted to a common type according to the rules of C99 or if two arguments were not provided.% (an exercise for the reader: modify this method to also allow a single argument for when the imaginary part is 0).
+
+As seen in Listing \ref{mapout}, we are implementing complex numbers internally using OpenCL vector types, like \verb|int2|.
+Let us look first at \verb|trans_New| on lines \ref{cplx}.54-\ref{cplx}.60, where new complex numbers are translated to vector literals by invoking \verb|target.VecLit|.
+This will ultimately  generate the necessary OpenCL code, as a string, to complete compilation (these strings are not directly manipulated by extensions, however, to avoid problems with, e.g.
+precedence).
+For it to be possible to reason compositionally about the correctness of compilation, all complex numbers must translate to terms in the target language that have a consistent target type.
+The \verb|trans_type| method of the \verb|ace.ActiveType| associates a type in the target language, here a vector type like \verb|int2|, with the active type.
+Ace supports a mode where this \emph{representational consistency} is dynamically checked during compilation (requiring that the active target know how to assign types to terms in the target language, which can be done  for our OpenCL target as of this writing).
+
+The translation methods for attributes (line \ref{cplx}.12) and binary operations  (line \ref{cplx}.31) proceed in a straightforward manner.
+The context provides a method, \verb|trans|, for recursively determining the translation of subexpressions as needed.
+Of note is that the translation methods can assume that typechecking succeeded.
+For example, the implementation of \verb|trans_Attribute| assumes that if \verb|node.attr| is not \verb|'ni'| then it must have been \verb|'i'| on line \ref{cplx}.14, consistent with the implementation of \verb|type_Attribute| above it.
+Typechecking and translation are separated into two methods to emphasize that typechecking is not target-dependent, and to allow for more advanced uses, like type refinements and hypothetical typing judgements, that we do not describe here.
+
+\begin{codelisting}
+\lstinputlisting{listing9.py}
+\caption{\texttt{[}in \texttt{examples/clx.py]} The active type family \texttt{Ptr} implements the semantics of OpenCL pointer types.}
+\label{cplx}
+\end{codelisting}
+\section{Expressiveness}\label{examples}
+Thus far, we have focused mainly on the OpenCL target and shown examples of fairly low-level active types: those that implement OpenCL's primitives (e.g.
+\verb|clx.Ptr|), extend them in simple but convenient ways (e.g.
+\verb|clx.Cplx|) and those that make interactive execution across language boundaries safe and convenient (e.g.
+\verb|clx.NPArray|).
+Ace was first conceived to answer the question: \emph{can we build a statically-typed language with the semantics of a C but the syntax and ease-of-use of a Python?} We submit that the work described above has answered this question in the affirmative.
+
+But Ace has proven useful for more than low-level tasks like programming a GPU with OpenCL.
+We now describe several interesting extensions that implement the semantics of primitives drawn from a range of different language paradigms, to justify our claim that these mechanisms are highly expressive.
+
+\subsection{Growing a Statically-Typed Python Inside an Ace}
+Ace comes with a target, base and type implementing Python itself: \verb|ace.python.python|, \verb|ace.python.base| and \verb|ace.python.dyn|.
+These can be supplemented by additional active types and used as the foundation for writing actively-typed Python functions.
+These functions can either be compiled ahead-of-time to an untyped Python file for execution, or be immediately executed with just-in-time compilation, just like the OpenCL examples we have shown.
+Many of the examples in the next section support this target in addition to the OpenCL target we have focused on thus far.
+
+\subsection{Recursive Labeled Sums}
+\begin{codelisting}
+\lstinputlisting{datatypes_t.py}
+\caption{\texttt{[datatypes\_t.py]} An example using statically-typed functional datatypes.}
+\label{datatypest}
+\end{codelisting}
+\begin{codelisting}
+\lstinputlisting{datatypes.py}
+\caption{\texttt{[datatypes.py]} The dynamically-typed Python code generated by running \texttt{acec datatypes\_t.py}.}
+\label{datatypes}
+\end{codelisting}
+Listing \ref{datatypest} shows an example of the use of statically-typed functional datatypes (a.k.a.
+recursive labeled sum types) together with the Python implementation just described.
+It shows two syntactic conveniences that were not mentioned previously: 1) if no target is provided to \verb|ace.fn|, the base can provide a default target (here, \verb|ace.python.python|); 2) a concrete function can be generated immediately by providing a type assignment after \verb|ace.fn| using braces.
+Listing \ref{datatypest} generates Listing \ref{datatypes}.
+
+Lines \ref{datatypest}.3-\ref{datatypest}.11 define a function that generates a recursive algebraic datatype representing a tree given a name and another Ace type for the data at the leaves.
+This type implemented by the active type family \verb|fp.Datatype|.
+A name for the datatype and the case names and types are provided programmatically on lines \ref{datatypest}.3-\ref{datatypest}.8.
+To support recursive datatypes, the case names are enclosed within a lambda term that will be passed a reference to the datatype itself.
+These lines also show two more active type families: units (the type containing just one value), and immutable pairs.
+Line \ref{datatypest}.13 calls this function with the Ace type for dynamic Python values, \verb|py.dyn|, to generate a type, aliased \verb|DT|.
+This type is implemented using class inheritance when the target is Python, as seen on lines \ref{datatypes}.1-\ref{datatypes}.13.
+For C-like targets, a \verb|union| type can be used (not shown).
+
+The generic function \verb|depth_gt_2| demonstrates two features.
+First, the base has been setup to treat the final expression in a top-level branch as the return value of the function, consistent with typical functional languages.
+Second, the \verb|case| ``method'' on line \ref{datatypest}.17 creatively reuses the syntax for dictionary literals to express a nested pattern matching construct.
+The patterns to the left of the colons are not treated as expressions (that is, a type and translation is never recursively assigned to them).
+Instead, the active type implements the standard algorithm for ensuring that the cases cover all possibilities and are not redundant \cite{pfpl}.
+If the final case were omitted, for example, the algorithm would statically indicate an error, just as in statically-typed functional languages like ML.
+
+\subsection{Nested Data Parallelism}
+Functional constructs have shown promise as the basis for a nested data-parallel parallel programming model.
+Many powerful parallel algorithms can be specified by composing primitives like \verb|map| and \verb|reduce| on persistent data structures like lists, trees and records.
+Data dependencies are directly encoded in these specifications, and automatic code generation techniques have shown promise in using this information to automatically execute these specifications on concurrent hardware and networks.
+The Copperhead programming language, for example, is based in Python as well and allows users to express functional programs over lists using common functional primitives, compiling them to CUDA code \cite{catanzaro2011copperhead}.
+By combining active types and the metaprogramming facilities described in Sec.
+2 to implement optimizations, like fusion, on the typed syntax trees available from concrete functions, these same techniques can be implemented as an Ace library as well.
+
+\subsection{Product Types}
+Product types (types that group heterogeneous values together) like structs, unions, tuples, records and objects can all be implemented as Ace type families  parameterized by a dictionary mapping field names (indices, in the case of tuples) to types.
+Each family differs slightly in the semantics of the operations available.
+Structs support mutation, for example, while records do not.
+Object systems typically introduce additional logic for calling methods, binding special variables like \verb|this|, and accessing information through an inheritance hierarchy.
+Listing \ref{ooclxfppy} shows an example of each of these.
+The prototypic object system delegates to a backing record if a field is not available in the foreground.
+This example also demonstrates cross-compilation to C99, supported for a subset of operations in \verb|clx|.
+All three abstractions are implemented as simple \verb|struct|s, as can be seen in Listing \ref{ooclxfpc}.
+This example further demonstrates the use of argument annotations to generate concrete functions (line \ref{ooclxfppy}.9), which is a feature only available when using Python 3.3+.
+
+\subsection{Distributed Programming}
+Many distributed programming abstractions can be understood as implementing object models with complex forms of dynamic dispatch.
+For example, in Charm++, dynamic dispatch involves message passing over a dynamically load-balanced network \cite{kale2009charm++}.
+While Charm++ is a sophisticated system, and we do not anticipate that implementing it as an Ace extension would be easy, it is possible to do so by targeting a system like Adaptive MPI, which exposes the Charm++ runtime system \cite{kale2009charm++}.
+%Erlang and other message-passing and actor-based systems can also be considered in this manner -- processes can be thought of as objects, channels as methods and messages as calls.
+
+%\subparagraph{Annotation and Extension Inference}
+%In addition to type annotations, OpenCL normally asks for additional annotations in a number of other situations.
+Users can annotate functions that meet certain requirements  with the \verb|__kernel| attribute, indicating that they are callable from the host.
+The \verb|OpenCL| backend can check these and add this annotation automatically.
+Several types (notably, \verb|double|) and specialized primitive functions require that an OpenCL extension be enabled via a \verb|#pragma|.
+The OpenCL backend automatically detects many of these cases as well, adding the appropriate \verb|#pragma| declaration.
+An example of this can be seen in Listing \ref{mapout}, where the use of the \verb|double| type triggers the insertion of the \verb|cl_khr_f64| extension.
+%
+
+A number of recent languages designed for distributed programming on clusters use a partitioned global address space model, e.g.
+Chapel \cite{chapel}.
+These languages provide first-class support for accessing data transparently across a massively parallel cluster.
+Their type systems track information about \emph{data locality} so that the compiler can emit more efficient code.
+The extension mechanism can also track this information in a manner analogous to how address space information is tracked in the OpenCL example above, and target an existing portable run-time such as GASNet \cite{bonachea2002gasnet}.
+
+\subsubsection{Units of Measure}
+A number of domain-specific type systems can be implemented within Ace as well.
+For example, prior work has considered tracking units of measure (e.g.
+grams) statically to validate scientific code \cite{conf/cefp/Kennedy09}.
+This cannot easily be implemented using many existing abstraction mechanisms because this information should only be maintained statically to avoid excessive run-time overhead associated with tagging/boxing, and the typechecking logic is reasonably complex.
+The Ace extension mechanism allows this information to be tracked in the type system, but not included during translation, by a mechanism similar to how address space information is tracked with pointers.
+Because Ace extensions can use Python, the needed logic can be implemented directly.
+
+\subsection{OOFPCLX}
+\begin{codelisting}[t]
+\lstinputlisting{ooclxfp.py}
+\caption{\texttt{[ooclxfp.py]} An example combining structs and immutable records using a prototype-based object system, cross-compiled to C99. Uses Python 3 argument annotations.}
+\label{ooclxfppy}
+\end{codelisting}
+\begin{codelisting}[t]
+\lstinputlisting[style=OpenCL]{ooclxfp.c}
+\caption{\texttt{[ooclxfp.c]} The C99 code generated by running \texttt{acec ooclxfp.py}.}
+\label{ooclxfpc}
+\end{codelisting}
+
 \end{document}

### 2014-03-24 ace-oopsla14/ace-oopsla14.tex
OOPSLA progress

@@
 First, they are themselves often available only within a dialect of an existing language and thus face a ``chicken-and-egg'' problem: a language like SugarJ \cite{erdweg2011sugarj} must overcome the same extrinsic issues as a language like HJ.
-Second, giving providers too much control over the language can lead to intrinsic safety issues, ambiguities and conflicts between extensions, as we will discuss.
-Too little control, on the other hand, leaves it difficult to implement interesting abstractions.%It is difficult to guarantee that extensions are sound in isolation and more so to guarantee that they can be {safely} combined.
+Second, giving abstraction providers too much control over the language (by introducing an over-zealous ``solution'' to the \emph{expression problem}) can introduce type safety issues and ambiguities that only become apparent when extensions are combined, as we will discuss.
+Too little control, on the other hand, leaves it difficult to implement real-world abstractions.%It is difficult to guarantee that extensions are sound in isolation and more so to guarantee that they can be {safely} combined.
 %This requires further research before it would be appropriate to widely rely on them (which we will discuss as we go on).
 
-This paper describes the design and implementation of Ace, an internally extensible, statically typed programming language designed considering both extrinsic and intrinsic factors.
-To address the ``chicken-and-egg'' problem, Ace is implemented within Python \cite{Politz:2013:PFM:2509136.2509536,python}, entirely as a library.
-The top-level of an Ace file is a \emph{compilation script} written in Python and Python serves as Ace's \emph{type-level language}, as we will discuss.
-Ace and Python share a common syntax and package system, so users of Ace can directly leverage established tools, infrastructure and coding standards.
-%The semantics of  define run-time behavior.
-
-Functions marked by a decorator as under the control of Ace are, unlike Python functions, \emph{statically} typechecked.
-This makes it possible to rule out many potential issues ahead of execution, and thus  avoid  costly dynamic checks.
-% Instead, Ace has a statically-typed semantics that can be extended by users from within libraries.
-However, there is \emph{no particular type system built into Ace}.
-Instead, a type is any  object implementing the \verb|ace.Type| interface constructed by the compilation script.
-We call these \emph{active types}.
-When statically assigning a type to a compound form, e.g.
-\verb|e.x|, the Ace {compiler} delegates to the type of a subexpression, e.g.
-to the type of \verb|e|, to determine how to assign a type to the expression as a whole.
-As we will show, this {type-directed protocol}, \emph{active typing}, permits the expression of a rich variety of type systems as libraries.
-Unlike syntax-directed protocols, there cannot be ambiguities when separately defined type sytems are combined (types control non-overlapping sets of expressions by construction).
-Base cases, e.g.
-variables and certain statement forms, are handled on a {per-function basis} by a \emph{base semantics}, also a compile-time object implementing an interface, \verb|ace.Base|.
-
-% As a result, clients are able to import any combination of extensions with the confidence that link-time ambiguities cannot occur .
-
-The dynamic behavior of an Ace function is determined by translation to a \emph{target language}.
-We begin by  targeting Python, then discuss targeting OpenCL and CUDA, lower-level languages used to program many-core processors (e.g.
-GPUs).
-Active types and bases control translation by the same type-directed protocol that governs type assignment.
-We call this phase \emph{active translation}.
-A \emph{target} mediates this process and is also a user-defined compile-time object implementing an interface, \verb|ace.Target|.
-%One base or type can support multiple targets. 
-When targeting a typed language, care must be taken to ensure that well-typed Ace expressions have well-typed translations, so we integrate a technique for compositionally reasoning about compiler correctness (developed for the TIL compiler for Standard ML \cite{tarditi+:til-OLD}) into the language.
-
-Ace can be used from the shell, producing source files that can be further compiled or executed by external means.
+In this paper, we describe an extensible statically-typed language, Ace, that is implemented entirely as a library within Python, avoiding the ``chicken-and-egg problem'' and occupying what we intend to convey by example as a ``sweet spot'' in the design space (contribution 1, Sec.
+\ref{usage}).
+
+Ace, together with a core lambda calculus, serves also as a vehicle for introducing a novel extension mechanism we call \emph{active typechecking and translation}.
+We sidestep the expression problem entirely by leaving the forms of expressions fixed, sharing them with Python.
+Instead, abstraction providers extend the language by introducing new type constructors and operators using type-level functions.
+We show how, by repurposing the conventional syntactic forms found in languages like Python, structuring the language as a bidirectional type system and integrating techniques from the typed compilation literature directly into the language, we can maintain expressiveness while  guaranteeing that link-time ambiguities cannot occur and type safety is maintained (contribution 2, Secs.
+\ref{att} and \ref{theory}).
+
+%Ace can be used from the shell, producing source files that can be further compiled or executed by external means.
 Ace functions targeting a language with Python bindings (e.g.
 Python itself, as well as the others just described) can also be compiled and invoked interactively from Python.
-Ace supports a form of \emph{ad hoc polymorphism} (based on singleton types) whereby types propagate into generic functions.
-This occurs both between Ace functions and when calling Ace functions from Python.
-In the latter case, the ``dynamic type'' of a Python value determines a static type ``just-in-time'' when a call occurs.
-This represents a form of implicit \emph{staging} and significantly streamlines interactive workflows (like those in scientific computing) that alternate between Python for orchestration and exploratory tasks (e.g.
-plotting) and an external language via a foreign-function interface (FFI) for performance- and correctness-critical portions.%We will discuss in detail as our primary realistic case study.
-
-The remainder of the paper is organized as follows: in Sec.
+Ace supports a form of \emph{ad hoc} polymorphism (based on singleton types) whereby the ``dynamic type'' of a Python value can determine a static type ``just-in-time'' when Python code calls a generic Ace function.
+This represents a form of implicit \emph{staging} and we show how this can significantly streamline interactive workflows (like those in scientific computing) that alternate between Python for orchestration and exploratory tasks (e.g.
+plotting) and a statically-typed target  language via a foreign-function interface (FFI) for performance- and correctness-critical portions.
+We discuss a complete implementation of the OpenCL type system (a language for low-level data parallel programming on GPUs), show several higher-level extensions to it, and discuss deep bindings between Ace, the OpenCL run-time system and Python's standard numerics package, \verb|numpy|, all implemented within a library (contribution 3, Sec.
+\ref{targets}).%We will discuss in detail as our primary realistic case study.
+
+%The remainder of the paper is organized as follows: in Sec.
 \ref{usage}, we describe the basic structure and usage of Ace with an example that uses several interesting Ace type libraries.
@@
 \ref{targets}, we discuss generic functions, FFIs and staging via a complete implementation of the OpenCL type system (and extensions to it) as a library.
+
 In Sec.
-\ref{related}, we compare Ace to related work.
+\ref{related}, we compare Ace to related work on language extensibility, term rewriting and staged compilation.
 We conclude in Sec.
-\ref{discussion} by summarizing our contributions, highlighting the key features needed by a host language to support these mechanisms, and discussing limitations and potential future work.
-The appendix contains details omitted from the main body and several additional examples.
+\ref{discussion} by highlighting the key features needed by a host language to support active typechecking and translation, and discussing limitations and potential future work.
+The appendix contains details omitted for concision as well as several more examples, including functional datatypes, flexible complex numbers and a previously little-explored type system for tracking string invariants using regular expressions statically (which we show being used in Listing \ref{example}, amongst several other examples).
 
@@
 %
-Listing \ref{example} shows an example of an Ace compilation script.
+Listing \ref{example} is an example of an Ace \emph{compilation script}.
 As promised, compilation scripts are written directly in Python.
@@
 \verb|pip|) and package repostories (e.g.
-\verb|PyPI|) are directly available for distributing libraries that use Ace.
+\verb|PyPI|) are directly available for distributing Ace-based libraries.
 Note that all of the type systems that we discuss are implemented entirely as  libraries.
+%Ace is agnostic about the particular design choices that we made in our examples.
 
@@
 from examples.regex import string_in
+
+print "Hello, compile-time world!"
 
@@
 
-print "Hello, compile-time world!"
+print "Goodbye, compile-time world!"
 \end{lstlisting}
@@
 \subsection{Functions}
+
 Functions implementing run-time behavior (\emph{typed functions}) are distinguished by the presence of a decorator specifying their  \emph{base semantics} (here \verb|py| from \verb|examples.py|; we will discuss this further below).
@@
 Ace  leverages the \verb|inspect| and \verb|ast| modules in the Python standard library to extract abstract syntax trees for typed functions \cite{python}.
+
+On lines 14-18, we see the function \verb|log_transfer|.
+As in Python, it begins with a documentation string  that informally specifies the behavior of the function.
+Before moving into the body, however, we  also write a \emph{type signature} (line 17) stating that the type of \verb|t| is \verb|Transfer|, the prototypic object type described above.
+As a result, we can assume on line 19 that \verb|t.amount| and \verb|t.name| have types \verb|decimal[2]| and \verb|string|, respectively.
+We will return to the details of \verb|print| and the form \verb|[string](t.amount)|, which performs an explicit conversion, shortly.\todo{when?}
+
+Line 17 borrows Python's syntax for dictionary literals to approximate  conventional notation for type annotations.
+%\footnote{In the rare situation where there are no type annotations and the first statement is a dictionary literal, users can use Python's \texttt{pass} keyword.} % (and following conventions for documentation strings, which also describe ``types'' at the end).
+% The ``keyword'' \verb|type| was chosen because it is the name of a built-in function in Python (and thus would not be used in the way shown for any other purpose).
+ {In version 3.0 of Python, syntax for annotating arguments with arbitrary values directly was introduced \cite{pep3107}:
+% \vspace{-3px}
+\begin{lstlisting}[numbers=none]
+def log_transfer(t : Transfer): 
+  print ...
+\end{lstlisting}
+\vspace{-5px}
+These annotations were initially intended to serve only as documentation (suggesting that users of dynamically typed languages see potential in a typing discipline, if not a static one), but they were also made available as metadata for use by unspecified  future libraries.
+Ace supports both notations when the compilation script is run using Python 3.\textit{x}, but we use the more universally available notation here in view of our extrinsic goals: the Python 2.\text{x} series  remains the most widely used by a large margin as of this writing.% (though we consider cross-compilation in Sec.
+\ref{targets}).
+
+\subsection{Bidirectional Typechecking of Introductory Forms}
+On lines 21-30, we define the function \verb|__toplevel__| (the base will consider this a special name for the purposes of compilation, discussed in the Sec.
+\ref{compilation}).
+After printing a run-time greeting, we construct a \emph{value} of type \verb|C| on lines 24-26.
+Recall that \verb|C| is a record type, so we provide the names of the fields (here, without quotes because we are no longer in the type-level language) and values of the appropriate type using the syntactic form normally used for dictionary literals.
+We specify which record type the literal should be {analyzed} against by giving a \emph{literal ascription}, \verb|(C)|.
+This again repurposes existing  syntax: here function application.
+When the ``function'' is of literal form (dictionaries, tuples, lists, strings, numbers, booleans and  \verb|None|) and the argument is a type, we consider it instead an \emph{ascribed introductory form} of that type.
+We see another such form used with Python's tuple syntax on lines 27-28 to introduce terms of type \verb|Transfer| by providing the fore and prototype.
+
+%\subsection{Bidirectional Typechecking}
+The string, number and dictionary literal forms inside the outermost forms on lines 24-28 are also  introductory, but do not have an ascription.
+This is because the definition of the type \verb|C| completely determines which types they will need to have.
+As we will discuss in more detail shortly, Ace is built around a \emph{bidirectional type system} that distinguishes locations where an expression must \emph{synthesize} a type (e.g.
+to the right of bindings) from those where it can be \emph{analyzed} against a known type (e.g.
+as an argument to a function with known type) \cite{Lovas08abidirectional}.
+Literal forms are only analytic.
+
+If a literal is used in a synthetic location, the base can, however, specify a ``default ascription''.
+For example, if we had not specified the literal ascription on line 26, the base would cause the dictionary literal to be analyzed against the type \verb|dyn|, covering dynamically classified Python values.
+This would still lead to a type error because the variables used as keys are not bound in the top-level context (emphasizing that ``dynamically-typed languages'' can still have a static semantics, even if it only involves checking that top-level variables are bound).% When considered as a record, the keys were considered in a different class by the implementation (labels, rather than variables).
+
+An ascription can also evaluate statically to a {type constructor}.
+For example, we might write \verb|[1, 2] (matrix)| instead of \verb|[1, 2] (matrix[i32])| when using a base for which the default ascription for number literals is \verb|i32|.
+Because the arguments are synthetic, the type constructor can determine an appropriate index based on the types of the subexpressions.
+If we wanted a matrix of \verb|f32|s, then we would have to write either \verb|[1, 2] (matrix[f32])| or \verb|[1(f32), 2(f32)] (matrix)|, or use a base with a different default ascription.
+We will see how bidirectional typechecking of introductory forms  works in greater technical detail in Secs.
+\ref{att} and \ref{theory}, but note here that it is one of the key mechanisms for expressiveness given our constraint that we cannot add entirely new forms to the language (and indeed, doing so would begin to be quite crowded even if we could add distinct forms for matrices, vectors, lists, tuples, etc.).
+
+\subsection{External Compilation}\label{compilation} 
 \begin{codelisting}
@@
 Hello, compile-time world!
+Goodbye, compile-time world!
 [acec] _listing1.py successfully generated.
@@
 t2 = ((15,0), common)
-print_check(t1)
-print_check(t2)
+log_transfer(t1)
+log_transfer(t2)
 \end{lstlisting}
@@
 
-On lines 14-18, we see the function \verb|log_transfer|.
-As in Python, we write a documentation string  that informally specifies the behavior of the function.
-Before moving into the body, however, we  also write a \emph{type signature} (line 17) stating that the type of \verb|t| is \verb|Transfer|, the prototypic object type described above.
-As a result, we can assume on line 19 that \verb|t.amount| and \verb|t.name| have types \verb|decimal[2]| and \verb|string|, respectively.
-We will return to the details of \verb|print| and the form \verb|[string](t.amount)|, which performs an explicit conversion, shortly.\todo{when?}
-
-Line 17 borrows Python's syntax for dictionary literals to approximate  conventional notation for type annotations.
-%\footnote{In the rare situation where there are no type annotations and the first statement is a dictionary literal, users can use Python's \texttt{pass} keyword.} % (and following conventions for documentation strings, which also describe ``types'' at the end).
-% The ``keyword'' \verb|type| was chosen because it is the name of a built-in function in Python (and thus would not be used in the way shown for any other purpose).
- {In version 3.0 of Python, syntax for annotating arguments with arbitrary values directly was introduced \cite{pep3107}:
-% \vspace{-5px}
-\begin{lstlisting}[numbers=none]
-def log_transfer(t : Transfer): ...
-\end{lstlisting}
-\vspace{-5px}
-These annotations were initially intended to serve only as documentation (suggesting that users of dynamically typed languages see potential in a typing discipline, if not a static one), but they were also made available as metadata for use by unspecified  future libraries.
-Ace supports both notations when the compilation script is run using Python 3.\textit{x}, but we use the more universally available notation here in view of our extrinsic goals: the Python 2.\text{x} series  remains the most widely used by a large margin as of this writing (though we consider cross-compilation in Sec.
-\ref{targets}).
-
-\subsection{Bidirectional Typechecking of Introductory Forms}
-On lines 21-30, we define the function \verb|__toplevel__| (the base will consider this a special name for the purposes of compilation, discussed in the Sec.
-\ref{compilation}).
-After printing a run-time greeting, we construct a \emph{value} of type \verb|C| on lines 24-26.
-Recall that \verb|C| is a record type, so we provide the names of the fields (here, without quotes because we are no longer in the type-level language) and values of the appropriate type using the syntactic form normally used for dictionary literals.
-We specify which record type the literal should be {analyzed} against by giving a \emph{literal ascription}, \verb|(C)|.
-This again repurposes existing  syntax: here function application.
-When the ``function'' is of literal form (dictionaries, tuples, lists, strings, numbers and keywords like \verb|True| and \verb|None|) and the argument is a type, we consider it instead an \emph{ascribed introductory form} of that type.
-We see the same form used with Python's tuple syntax on lines 27-28 to introduce terms of type \verb|Transfer| by providing the fore and prototype.
-
-%\subsection{Bidirectional Typechecking}
-The string, number and dictionary forms inside the outermost forms on lines 24-28 are also introductory forms, but do not require an ascription.
-This is because the definition of the type \verb|C| already determines which types they should be analyzed.
-As we will discuss in more detail shortly, Ace is built around a \emph{bidirectional type system} that distinguishes locations where an expression must \emph{synthesize} a type (e.g.
-to the right of bindings) from those where it can be \emph{analyzed} against a known type (e.g.
-as an argument to a function with known type) \cite{Lovas08abidirectional}.
-Literal forms can only be analyzed against a type.
-If used in a synthetic location, the base must specify a ``default ascription''.
-For example, if we had not specified the literal ascription on line 26, the base would cause it to be analyzed against a type for Python dictionaries, which would lead to a type error because the variables used as keys are not bound in the surrounding context.
-
-An ascription can also be a {type constructor}, rather than a type.
-For example, we might write \verb|[1, 2] (matrix)| rather than \verb|[1, 2] (matrix[i32])| assuming the default ascription for integer literals is \verb|i32|.
-The type constructor synthesizes an index based on the types synthesized by the subexpressions.
-If we wanted a matrix of \verb|f32|s, then we would have to write either \verb|[1, 2] (matrix[f32])| or \verb|[1(f32), 2(f32)] (matrix)|, or use a base with a different default ascription.
-We will see how this works in greater technical detail shortly.
-
-\subsection{External Compilation}\label{compilation} 
 To typecheck, translate and execute the code in Listing \ref{example}, we have two choices: do so externally at the shell, or perform compilation interactively from within Python.
-We will here describe the former, then return to interactive compilation in Sec.
-\ref{targets} with a different example.
+We will begin with the former, then return to the latter in Sec.
+\ref{targets}.% with our OpenCL examples.
 
 Listing \ref{external-compilation} shows how to invoke the \verb|acec| compiler at the shell to typecheck and translate \texttt{listing\ref{example}.py}, resulting in a file named \verb|_listing1.py|.
-This is then sent to the  Python interpreter to produce the desired run-time behavior.
-Note that the {\texttt{print}} statement on line 32 of the compilation script was  evaluated during compilation only.
-These two steps can be combined by running \verb|ace listing1.py| (the intermediate file is not generated in this case).
+This is then sent to the  Python interpreter for execution.
+Note that the {\texttt{print}} statements at the top-level of the  compilation script were evaluated during compilation only.
+These two steps can be combined by running \verb|ace listing1.py| (the intermediate file is not generated unless explicitly requested in this case).
 
@@
 \begin{enumerate}
-\item Executes the provided compilation script (\verb|listing1.py|)
-\item For any top-level bindings that are Ace functions in the final environment (instances of \verb|ace.TypedFn|, as we will discuss), it initiates the active type-checking and translation process described in Sec.
-\ref{att}.
-If no type errors are discovered, the translations are extracted and emitted (the file extension(s) are determined by the target(s) in use, discussed in Sec.
-\ref{targets}; here our target is  the default target associated with the base \verb|examples.py.py|, which emits Python 2.6 files).
+\item Evaluate the provided compilation script (\verb|listing1.py|)
+\item For any top-level bindings that are Ace functions in the final environment (instances of \verb|ace.TypedFn|, as we will discuss), it initiates active type-checking and translation (Sec.
+\ref{att}).
+If no type errors are discovered, the translations are collected (obeying order dependencies) and emitted, with file extension(s) determined by the target(s) in use, discussed in Sec.
+\ref{targets}.
+Here our target is  the default target associated with the base \verb|examples.py.py|, which emits Python 2.6 files.
 If a type error is discovered, no file is emitted and the error is displayed on the console.
@@
 \begin{enumerate}
-\item The base, recognizing the function name \verb|__toplevel__| as special, placed the translation of its body  at the top level of the file.
+\item The base recognized the function name \verb|__toplevel__| as special, placing the translation of its body  at the top level of the file.
 Ace does not have any special function names itself; this is a feature of the chosen base.
-\item Records with two or more fields translated to tuples of their values. If there was only a single field, like \verb|A|, its value was passed around unadorned. 
-\item Decimals are represented as pairs of integers.
-Conversion to a string happens via a helper function defined in a ``runtime'' package imported with an internal name, \verb|__ace_0|, to avoid naming conflicts.
+\item Records with two or more fields translated to tuples of their values (e.g.
+\verb|common| on line 8).
+If there was only a single field, like the terms of type \verb|A| inside \verb|t1| and \verb|t2|, the value was passed around unadorned.
+\item Decimals translated to pairs of integers.
+Conversion to a string happened via a helper function defined in a ``runtime'' package imported with an internal name, \verb|__ace_0|, to avoid naming conflicts.
 \item Terms of type \verb|string_in[r"..."]| translated to strings.
-Checks for well-formedness were performed entirely statically by the type system.\todo{Nathan?}\footnote{Note that there are situations (e.g.
+Checks for membership in the specified regular language were performed entirely statically by the type system.\todo{Nathan?}\footnote{Note that there are situations (e.g.
 if a string is read in from the console) that would necessitate an initial run-time check, but no further checks in downstream functions would be necessary.
@@
 \subsection{Type Errors}
-Listing \ref{oops} shows an example of code containing several type errors.
-Indeed, lines 9-13 each contain a type error.
-If analagous code were written in Python itself, these could only be found if the code was executed on the first day of the month (and, depending on the implementation, not all of the issues would  produce even run-time exceptions).
-Static type checking allows us to find these errors during compilation.
-Listing \ref{oops-compilation} shows the result of attempting to compile this code.
-The compilation script completes (so that functions can refer to each other mutually recursively), then the typed functions in the top-level environment are typechecked.
-The typechecker raises an exception statically at the first error, and \verb|acec| prints it as shown.
 \begin{codelisting}[t]
@@
 \label{oops-compilation}
-\end{codelisting}
+\end{codelisting}Listing \ref{oops} shows an example of code containing several type errors.
+Indeed, lines 9-13 each contain a type error.
+If analagous code were written in Python itself, these could only be found if the code was executed on the first day of the month (and, depending on the implementation, not all of the issues would  immediately result in run-time exceptions, possibly leading to quite subtle problems).
+Static type checking allows us to find these errors during compilation.
+Listing \ref{oops-compilation} shows the result of attempting to compile this code.
+The compilation script completes (so that functions can refer to each other mutually recursively), then the typed functions in the top-level environment are typechecked.
+The typechecker raises an exception statically at the first error, and \verb|acec| prints it as shown.
 
@@
 
-In tooling for a monolithic language, there is typically a \emph{syntax-directed} protocol governing type synthesis/analysis (collectively, \emph{typechecking}\todo{rename it above}), translation, pretty-printing and so on.
-For example, one might declare functional datatypes that capture all types and term forms, and then proceed by  exhaustive case analysis to implement these functions.
+In compilers and other tools for monolithic languages, there is typically a \emph{syntax-directed} protocol governing type checking, translation, pretty-printing and so on.
+For example, one might declare functional datatypes that capture all types and term forms, and then proceed by  exhaustive case analysis over terms to implement this functionality.
 The external visitor pattern for class-based object systems captures essentially the same protocol.
-In both cases, the problem is that all the expression forms and types are defined in one place, so one cannot extend the language modularly.
-
-In extensible languages the focus is typically on enabling the addition of new types and term forms.
+In both cases, the problem is that the types and terms are defined in one place, so  providers cannot extend the language modularly.
+
+In extensible languages the focus is typically on enabling the addition of both new types and term forms.
 Adding new term forms requires then adding cases for all possible functions over terms (e.g.
 typing, translation,  code highlighting, style checking and so on).
-This is difficult (we will discuss some approaches in Sec.
+This is difficult.
+Indeed, this is the canonical example of the widely-studied \emph{expression problem} \cite{wadler1998expression}.
+Although there are some approaches that work for certain purposes, retaining the ability to reason compositionally about metatheoretic issues (e.g.
+type safety) and avoiding ambiguities when extensions are combined can become difficult (see Sec.
 \ref{related}).
-Doing so while retaining the ability to reason compositionally about metatheoretic issues (e.g.
-type safety) is even more difficult.
-This is the canonical example of the widely-studied \emph{expression problem} \cite{wadler1998expression}.
-
-Ace's term forms, as we have seen, are fixed by Python's grammar, so Ace sidesteps the expression problem entirely (and thus reaps the extrinsic benefits related to tool support, described earlier).
-Instead of taking a syntax-directed view of extensibility, we take a type-directed view: users cannot add new forms but they can add new types, which can affect the semantics of existing forms by an unambiguous protocol.
+%Requiring that abstraction clients have the expertise to resolve these ambiguities manually is undesirable, and heading them off results in an unbounded, combinatoric explosion of situations to consider.
+
+Ace's term forms, as we have seen, are fixed by Python's grammar, so Ace sidesteps the expression problem almost entirely (and thus inherits broad tool support, as described earlier).
+The key insight is that leaving the forms fixed does not mean the semantics must also be fixed.
+Instead of taking a syntax-directed view of extensibility, we take a type-directed view: users cannot add new forms but they can add new types, which we will allow to affect the semantics of or reinterpret existing forms (e.g.
+literals, as discussed above, but also nearly every other form, as we will now discuss) in a controlled manner.
 %In Ace (but not in the theory in Sec.
 \ref{theory}), users can also repurpose existing forms flexibly within a delimited scope.
-More specifically, the key mechanisms that distinguish Ace's approach to extensibility are: \begin{itemize}
+More specifically, the key features that distinguish active typechecking and translation are: \begin{itemize}
 \item a \emph{dispatch protocol} that delegates responsibility for typechecking and translation to:
   \begin{itemize}
-  \item a type or type constructor extracted from a subterm of an expression in some way
+  \item a type or type constructor extracted from a subterm of a compound expression
   \item a per-function base semantics (or simply \emph{base}) for base forms for which this is not possible  (statements, variables and unascribed literals)
@@
 \subsection{Base Decorators}
-Function decorators in Python (line 2) are syntactic sugar for \emph{calling} the decorator (\verb|py|) with the decorated function (lines 3-5) as an argument, then replacing the binding (\verb|example|)  with the return value of this call.
+Function decorators in Python (line 2) are syntactic sugar for {calling} the decorator (\verb|py|) with the decorated function (lines 3-5) as an argument, then replacing the binding (\verb|example|)  with the return value of this call.
 We could equivalently have omitted line 3 and inserted the top-level statement \verb|example = py(example)| on the line after the function definition.
-Note that \verb|py| is not a function but an instance of a class, \verb|examples.py.PyBase|, inheriting from \verb|ace.Base|.
-It can be used as a decorator because \verb|ace.Base| overloads the call operator as shown in Listing \ref{base}.
+Note, however, that \verb|py| is not a function but an instance of a class, \verb|examples.py.PyBase|, that inherits from \verb|ace.Base|.
+It can be used as a decorator because \verb|ace.Base| overloads the call operator, shown in Listing \ref{base}.
 The \verb|_aceify| function (not shown) \todo{show in appendix?}operates as follows:
@@
 \subsection{Typechecking Typed Functions and Statements}
+When the Ace compiler begins typechecking a \verb|TypedFn| (when asked to by \verb|acec|, or when it is  invoked interactively as we will show in Sec. \ref{targets}), it proceeds as follows:
+\begin{enumerate}
+\item An \verb|ace.Context| is constructed with a reference to the function (Listing \ref{context}, lines 2-3).
+\item The base is asked to initialize the context.
+This can be seen on lines 5-7 of Listing \ref{pybase}, where the Python base adds an attribute tracking local bindings (initially only the arguments) and the return type to be synthesized (initially not known, indicated using \verb|None|).
+\item For each statement in the body (ignoring the documentation string and type annotation, if present), the compiler calls a method of the base, \verb|check_|$F$, where $F$ is the form of the statement (taken directly from Python's \verb|ast| package \cite{python}).
+These methods are responsible for checking that statement is well-typed (statements do not themselves have a type) and implementing any necessary effects on the context.
+
+The assignment statements on lines 5 and 6 of Listing \ref{att-example} trigger the method \verb|check_Assign| \todo{if time, add destructuring assignment to appendix}shown on lines 6-14 of Listing \ref{pybase} and the return statement on line 7 triggers the method \verb|check_Return| on lines 18-22.
+Both work similarly: if this is the first assignment to a particular identifier, or the first return statement seen, then the value must synthesize a type in the current context.
+Otherwise, it is analyzed against the previously synthesized type.
+\item The base synthesizes a type for the function as a whole via the \verb|syn_ty_FunctionDef_outer| method.
+Here, an \verb|arrow| type is synthesized (lines 26-29) based on the argument signature and the synthesized return type.
+\end{enumerate}
+
+\subsection{Bidirectionally Typechecking Expressions}
 \begin{codelisting}
@@
 \end{codelisting}
-When the Ace compiler begins typechecking a \verb|TypedFn| (when asked to by \verb|acec|, or when it is  invoked interactively as we will show in Sec. \ref{targets}), it proceeds as follows:
-\begin{enumerate}
-\item An \verb|ace.Context| is constructed with a reference to the function (Listing \ref{context} lines 2-3).
-\item The base is asked to initialize the context.
-This can be seen on lines 5-7 of Listing \ref{pybase}, where the Python base adds an attribute tracking local bindings (initially only the arguments) and the return type to be synthesized (initially not known, indicated using \verb|None|).
-\item For each statement in the remainder of the body, the compiler calls a method of the base, \verb|check_|$F$, where $F$ is the form of the statement (taken directly from Python's \verb|ast| package \cite{python}).
-These methods are responsible for checking that statement is well-typed (statements do not themselves have a type) and implementing any necessary effects on the context.
-\item The base synthesizes a type for the function as a whole via the \verb|syn_ty_FunctionDef_outer| method.
-\end{enumerate}
-
-The assignment statement on lines 5 and 6 of Listing \ref{att-example} triggers the method \verb|check_Assign| \todo{if time, add destructuring assignment to appendix}shown on lines 6-14 of Listing \ref{pybase} and the return statement on line 7 triggers the method \verb|check_Return| on lines 18-22.
-Both work similarly: if this is the first assignment to a particular identifier, or the first return statement seen, then the value must synthesize a type in the current context.
-Otherwise, the value is analyzed against the previously determined type.
-At the end, an \verb|arrow| type is synthesized (lines 26-29) based on the argument signature and the synthesized return type.
-
-\subsection{Bidirectionally Typechecking Expressions}
+
 \begin{codelisting}
@@
 \section{Discussion}\label{discussion}
+
+\begin{enumerate}
+\item We address the ``chicken-and-egg'' problem by designing Ace, an extensible language  embedded within Python \cite{Politz:2013:PFM:2509136.2509536,python} entirely as a library.
+The top-level of an Ace file is a \emph{compilation script} written in Python and Python serves as Ace's \emph{type-level language}, as we will discuss.
+Ace and Python thus share a common syntax and package system, so users of Ace can directly leverage established tools, infrastructure and coding standards.
+Functions marked by a decorator as under the control of Ace are, unlike Python functions, \emph{statically} typechecked.
+This makes it possible to rule out many potential issues ahead of execution, and thus  avoid  costly dynamic checks.
+%The semantics of  define run-time behavior.
+\item  % Instead, Ace has a statically-typed semantics that can be extended by users from within libraries. 
+There is \emph{no particular type system built into Ace}.
+Instead, a type is any  object implementing the \verb|ace.Type| interface constructed by the compilation script.
+We call these \emph{active types}.
+When statically assigning a type to a compound form, e.g.
+\verb|e.x|, the Ace {compiler} delegates to the type of a subexpression, e.g.
+to the type of \verb|e|, to determine how to assign a type to the expression as a whole.
+As we will show, this {type-directed protocol}, \emph{active typing}, permits the expression of a rich variety of type systems as libraries.
+Unlike syntax-directed protocols, there cannot be ambiguities when separately defined type sytems are combined (types control non-overlapping sets of expressions by construction).
+Base cases, e.g.
+variables and certain statement forms, are handled on a {per-function basis} by a \emph{base semantics}, also a compile-time object implementing an interface, \verb|ace.Base|.
+\end{enumerate}
+% As a result, clients are able to import any combination of extensions with the confidence that link-time ambiguities cannot occur .
+
+The dynamic behavior of an Ace function is determined by translation to a \emph{target language}.
+We begin by  targeting Python, then discuss targeting OpenCL and CUDA, lower-level languages used to program many-core processors (e.g.
+GPUs).
+Active types and bases control translation by the same type-directed protocol that governs type assignment.
+We call this phase \emph{active translation}.
+A \emph{target} mediates this process and is also a user-defined compile-time object implementing an interface, \verb|ace.Target|.
+%One base or type can support multiple targets. 
+When targeting a typed language, care must be taken to ensure that well-typed Ace expressions have well-typed translations, so we integrate a technique for compositionally reasoning about compiler correctness (developed for the TIL compiler for Standard ML \cite{tarditi+:til-OLD}) into the language.
+
+TODO
+
 Static type systems are powerful tools for programming language design and implementation.

### 2014-03-25 ace-oopsla14/string-appendix.tex
Moving work to cyrus-/papers and nuking clone.

\documentclass[10pt,preprint]{sigplanconf}
\usepackage{bussproofs}    % Gentzen-style deduction trees *with aligned sequents!*
\usepackage{mathpartir}	% For type-settting type checking rule figures
\usepackage{syntax}		% For type-setting formal grammars.

\usepackage{hyperref}		% For links in citations

\usepackage{float}			% Make figures float if [H] option is passed.

\iffalse
\usepackage{listings}		% For typesetting code listings
\usepackage{callout}
\usepackage{titlesec}
\usepackage[T1]{fontenc}
\usepackage{upquote}
\lstset{upquote=true}
\fi

\usepackage{textcomp}		% For \textquotesingle as used in introduction
\usepackage{color}			% for box colors, like in TAPL.

\usepackage{amsmath}		% Begin Carthage default packages
\usepackage{makeidx}
\usepackage{amsthm}
\usepackage{amsfonts}
\usepackage{amssymb}
\usepackage{graphics}
\usepackage{enumerate}
\usepackage{multicol}
\usepackage{epsfig}
\usepackage{csquotes}
\usepackage{enumitem}

\newtheorem{thm}{Theorem}                                                       
\newtheorem{cor}[thm]{Corollary}                                                
\newtheorem{lem}[thm]{Lemma}                                                    
\newtheorem{prop}[thm]{Proposition}                                             
\newtheorem{ax}[thm]{Axiom}                                                     
\theoremstyle{definition}                                                       
\newtheorem{defn}[thm]{Definition}                                              
\newtheorem{exam}[thm]{Example}                                                 
\newtheorem{rem}[thm]{Remark}

%
% For type setting inference rules with labels.
%
\newcommand{\inferlbl}[3]
			{\inferrule{#3}{#2}{\textsf{\footnotesize{\sc #1}}}}
\newcommand{\inferline}[3]
			{\inferrule{#3}{#2} & {\textsf{\footnotesize{\sc #1}}} \\ \\}

\newcommand{\Lagr}{\mathcal{L}}
\newcommand{\strin}[1]{{\tt string\_in #1}}
\newcommand{\lang}[1]{\Lagr\{#1\}}
\newcommand{\stru}[1]{ {\tt string\_union #1}}

\newcommand{\str}{ {\tt string} }
\newcommand{\istr}{ {\tt istring} }

\newcommand{\dconvert}{ {\tt dconvert} }
\newcommand{\filter}{ {\tt filter} }
\newcommand{\ifilter}{ {\tt ifilter} }

\newcommand{\reduces}{ \Rightarrow }
\newcommand{\ireduces}{ \leadsto }

\newcommand{\val}{ \ {\tt val} }
\newcommand{\ival}{ \ {\tt ival} }

\newcommand{\istrf}[1]{`#1\textrm'} %??
\newcommand{\strf}[1]{``#1"}

%
% Constrain the size of full-page diagrams and rule lists
%
%\newcommand{\pagewidth}{5in}
%\newcommand{\rulelistwidth}{3in}

%
% Names of type systems presented in paper
%
\newcommand{\lcs}{\lambda_{CS}}

\setlength{\grammarindent}{3em}

\begin{document}

TODO-nrf spell check.

TODO-nrf Include functions (app, abs)?

TODO-nrf Change icheck in eval rule to Python.

\section{Introduction}
In web applications and other settings, incorrect input sanitation often causes
security vulnerabilities. For this reason, frameworks often provide methods
for sanitizing user input. When these methods are insufficient or unavailable,
developers often write develop custom input sanitation algorithms. In both cases,
input sanitation techniques are ultimately implemented using the language's
regex library.

Formally, input sanitation is the problem of ensuring that an arbitrary string
is converted into a safe form before potentially unsafe use. For example, consider
SQL injection attacks. To prevent such attacks, we might ensure that any string
used as input to a query does not contain unescaped SQL common sequences.

This appendix presents a type system called $\lcs$ for ensuring that input 
sanitation algorithms are implemented correctly with respect to use site 
specifications.

Unlike frameworks provided by general purpose languages such as Haskell and
Ruby, our type system provides a \emph{static} guarantee that input is always 
properly sanitized before use. We achieve this by moving portions of the regular
expression library into the type system. Therefore, type safety relies upon 
several closure and decidability results about regular languages.

Although Ur/Web achieves a similar safety guarantee, its type system is significantly
more complicated than the system defined below. We believe this simplicity 
demonstrates the power of extensibility. Instead of introducing new abstractions
based upon subtle analyses and requiring the adoption of new programming languages, 
extension designers may capture natural programming idioms directly. TODO-nrf this seems unsubstantiated or dangerously vague.

Finally, XDuce typechecks XML. Like our work, XDuce relies upon some properties of
regular expression to establish soundness and copmleteness results. Unlike
XDuce, our work is motivated by input sanitation and therefore considers arbitrary
strings (as opposed to tree-structured XML documents). Furthermore, our static
treatment of the ubiquitous ``filter" function for regular expression is novel.

An outline of this appendix follows:

\begin{itemize}
  \item In \S 2, we define the type system's static and dynamic semantics.
  \item Section 3 recalls some classical results about regular expressions and presents a type safety proof for $\lcs$ based upon these properties.
  \item Finally, \S 4 discusses our implemention of $\lcs$ as a type system extension  within the Ace programming language.
\end{itemize}

\section{Definition of $\lcs$}

The $\lcs$ language is characterized by a type of strings indexed by regular
expressions, together with operations on such strings which correspond to common
input sanitation patterns.

This section presents the grammar and semantics of $\lcs$.
The semantics are defined in terms of an internal language with at least strings and a regex filter function.
These constraints are captured by the internal term valuations ($\ival$).
The internal language does not necessarily need a regex filter function because
any dynamic conversion is easily definable using a combination of filters and safe
casting.

%
% Begin Grammar 
%
\begin{figure}
\begin{grammar}
<r> ::= $\epsilon$ | $.$ | $a$ | $r \cdot r$ | $r + r$ | $r*$ \hfill Regex ($a \in \Sigma$).

<t> ::= 			\hfill terms: 					\\
%$x$					\hfill variable 				\alt
%${\tt lambda} x:t$		\hfill abstraction 			\alt
%$t\ t$				\hfill application				\alt
$\filter(\strin[r],t)$ 		\hfill filter substrings			\alt
$[\strin[r]](t)$ \hfill safe conversion \alt
$\dconvert[\strin[r]](t)$ \hfill unsafe conversion

<iv> ::= 			\hfill internal values:					\\
$\istrf{s}$				\hfill internal string \alt
${\tt ifilter}(r, \istr(s))$ \hfill internal filter

<v> ::= \hfill values: \\
$\strf{s}$ \hfill string

<T> ::=				\hfill	types:					\\
$\str$  \hfill Strings \alt
$\strin[r]$				\hfill Regular language strings		
%\alt
%$T \rightarrow T$	\hfill	Functions

<$\Gamma$> ::=	$\emptyset$	 | $\Gamma,x:T$ \hfill typing context
\end{grammar}
\caption{Syntax of $\lcs$}
\label{fig:lcsSyntax}
\end{figure}

The $\lcs$ language gives static semantics for common regular expression library
functions. In this treatment, we include concatenation and filtering.
The $\filter$ function removes all instances of a regular expression in a string,
while concatenation ($+$) concatenates two strings.

\onecolumn
\begin{figure}
\begin{center}
\begin{tabular}{c r}
%\inferline{T-VAR}{\Gamma \vdash x : T}{x:T \in \Gamma}
%
%\inferline{T-ABS}{\Gamma \vdash ({\tt \textbf{lambda}\ } x: t_2) : T_1 \rightarrow T_2}
%			{\Gamma, x:T_1 \vdash t_2:T_2}
%
%\inferline{T-APP}
%			{\Gamma \vdash t_1t_2 : T_{12}}
%			{\Gamma \vdash t_1 : T_{11}\rightarrow T_{12} \\ \Gamma \vdash t_2 : T_{11}}	
%
\inferline{T-Equiv-Top}
  {\str \equiv \strin[.*]}
  { }

\inferline{T-Equiv-string\_in}
  {\strin[r] \equiv \strin[r']}
  {\lang{r} = \lang{r'}}

\inferline{T-Equiv-Include}
  {\Gamma \vdash e:\strin[r]}
  {e \in \lang{r}}

\inferline{T-Minimum}
{ \Gamma \vdash e : \strin[e] }
{\Gamma \vdash e : \str}

%\inferline{T-Union}
%  {\Gamma \vdash \stru[r'](e) : \strin[r \cup r']}
%  { \Gamma \vdash e : \strin[r] }

\inferline{T-Concat}
  {\Gamma \vdash e_1 + e_2 : \strin[r_1 \cdot r_2]}
  { \Gamma \vdash e_1 : \strin[r_1] \\ \Gamma \vdash e_2 : \strin[r_2] }

\inferline{T-Convert}
  {\Gamma \vdash [\strin[r]](e):\strin[r]}
  {\Gamma \vdash e : \strin[r'] \\ \lang{r'} \subseteq \lang{r}}

\inferline{T-Dconvert (optional)}
{\Gamma \vdash \dconvert[\strin[r]](e):\strin[r]}
{\Gamma \vdash e : \strin[r']}

\inferline{T-Filter}
  {\Gamma \vdash \filter(\strin[r'], t) : \strin[(r \backslash r') + \emptyset]}
  {\Gamma \vdash t:\strin[r]}

\end{tabular}
\caption{Typing relation for $\lcs$}
\label{fig:lcsTyping}
\end{center}
\end{figure}

%
% Begin Operational Semantics
%
\begin{figure}
\begin{center}
\begin{tabular}{c r c r}
%\inferline{E-App1}{t_1t_2 \Rightarrow t_1't_2}{t_1 \Rightarrow t_2}
%\inferline{E-App2}{v_1t_2 \Rightarrow v_1t_2'}{t_2 \Rightarrow t_2'}
%\inferline{E-AppAbs}{({\tt lambda } x : t_{12})v_2 \Rightarrow [x / v_2]t_{12}}{T_{11}}

\inferline{E-strinval}
  {\strf{s}:\strin[r] \val}
  { }

%\inferline{E-strval}
%{\strf{s}:\str \val}
%  { }

\inferline{E-ival}
  {\istrf{e}:\istr \ival }
  { }

\inferline{E-ifilterval}
{ {\tt ifilter}(r,e) \ival }
{ e {\tt ival} }

\inferline{E-string}
{e : \str \ireduces \istrf{e}}
{ e \val }

\inferline{E-string\_in}
{e : \strin[r] \ireduces  \istrf{e}}
{ e \val }

\inferline{E-concat}
{e_1 + e_2 : \strin[r_1+r_2] \ireduces \istrf{e_1} + \istrf{e_2}}
{ }

\inferline{E-filterval}
{\filter(\strin[r], e) : \strin[r'] \reduces {\tt rl\_filter}(r,e)}
{ e \val }

\inferline{E-filter}
{\filter(\strin[r], e) : \strin[r'] \reduces  \filter(\strin[r], e' )}     
  { e : T \reduces e' }

\inferline{E-dconvertval (optional) \footnote{See definition ???} }
{\dconvert[\strin[r]](e) : \strin[r'] \reduces {\dconvert[\strin[r]](e'})}
{e:T \reduces e'}

\inferline{E-dconvert (optional) \footnote{See definition ???} }
{\dconvert[\strin[r]](e) : \strin[r'] \ireduces {\tt ifilter}(r, \istrf{e})}
{e \val }

\end{tabular}
\caption{Operational semantics for $\lcs$}
\label{fig:lcsEval}
\end{center}
\end{figure}
\twocolumn

\subsection{Typing}

The $\strin[r]$ type is parameterized by regular expressions; if $e:\strin[r]$,
then $e \in r$. Mapping from an arbitrary $\str$ to a $\strin[r]$ requires
defining a algorithm -- in terms of filter -- for converting a $\strin[.*]$
into a $\strin[r]$. The static semantics of the language defines the types of
operations on regular expressions in terms of well-understood properties about
regular lanugages; we recall these properties in section 3.

\subsection{Dynamics}

There are two evaluation judgements: $e:T \reduces e'$ and $e:T \ireduces i$.
The $\reduces$ relation is between $\lcs$ expressions, while the $\ireduces$
relation is a mapping from $\lcs$ expressions into internal language expressions
$i$ such that $i \ival$.

Safety of the evaluation relation depends upon an injective mapping from $\lcs$ types info
internal language types. This relation, $h$, is defined below.

\begin{defn}[Type translation]
  The type translation function $h : Type \rightarrow IType$ is defined as follows:
  \begin{itemize}
    \item $\forall r. h(\strin[r]) = \istr$
    \item $h(\str) = \istr$
  \end{itemize}
\end{defn}

\section{Type Safety}

The type safety proof relies upon some established results about regular languages.
The important definitions and theorems follow.

\subsection{Properties of Regular Languages}
 % copy from thesis

\subsection{Type Safety Proof}

\begin{thm}[Preservation]
  Let $T$ be a type in $\lcs$ and $h(T)=\sigma$ the corresponding type in the internal language.
  For all terms $e$:
  \begin{itemize}
    \item If $e:T$ and $e:T \ireduces i$ then $i : \sigma$ for some $\sigma$.
    \item If $e:T$ and $e:T \reduces e'$ then $e':T$.
  \end{itemize}
\end{thm}
\begin{proof}
The proof is a straightforward induction on the derivation of the combined evaluation relation.
%\begin{itemize}[label=$ $,itemsep=1ex]
%  \item \textbf{E-Ival, E-Ifilterval}. Both cases hold vacuously since the terms at hand are not $\lcs$ terms.
%  \item \textbf{E-String}. Since $e:\str$, it suffices to show that $h(\str) = \istr$ which holds by definition.
%  \item \textbf{E-Sring\_in}. Since $e:\strin[r]$, it suffices to show that $h(\strin[r]) = \istr$ which holds for any $r$ by definition.
%  \item \textbf{E-filter}. Note that ${\tt filter}(\strin[r], e) : \strin[r']$ for some $r'$. 
%    It suffices to show that rl\_filter(r,e)
%    $h(\strin[r]) = \istr$ which holds for any $r$ by definition.
%  \item \textbf{E-Dconvert}. Note that $\dconvert[r](e) : \strin[r]$, so it suffices to show that $h(\strin[r]) = \istr$ which holds by defintiion.
%\end{itemize}
\end{proof}

\begin{thm}[Progress]
  Let $T$ be a type in $\lcs$ and $\sigma$ a type in the internal language.
  For all terms $e$, if $e:T$ then $e:T \reduces^* e' \ireduces^* i$ where $i {\tt ival}$.
\end{thm}
\begin{proof}
By induction on the derivation of $e:T$.
%\begin{itemize}[label=$ $, itemsep=1ex]
%\end{itemize}
\end{proof}

\section{Implementation in Ace}

\end{document}

### 2014-03-25 ace-oopsla14/string-appendix.tex

@@
 \newcommand{\Lagr}{\mathcal{L}}
-\newcommand{\strin}[1]{{\tt string\_in #1}}
+\newcommand{\strin}{{\tt string\_in}}
 \newcommand{\lang}[1]{\Lagr\{#1\}}
-\newcommand{\stru}[1]{ {\tt string\_union #1}}
+\newcommand{\stru}[2]{ {\tt string\_union}(#1,#2)}
 
@@
 
-\newcommand{\dconvert}{ {\tt dconvert} }
-\newcommand{\filter}{ {\tt filter} }
-\newcommand{\ifilter}{ {\tt ifilter} }
+\newcommand{\dconvert}[2]{ {\tt dconvert}(#1,#2) }
+\newcommand{\filter}[2]{ {\tt filter}(#1,#2) }
+\newcommand{\ifilter}[2]{ {\tt ifilter}(#1,#2) }
 
@@
 TODO-nrf Change icheck in eval rule to Python.
+
+TODO-nrf Is it ok to make the proofs single-column? If not clean up overflows.
+
+TODO-nrf Make sure failure of unicity isn't causing any problems and write down lemmas about
+forms (e.g. $e \in \lang{r} \implies e = \strf{s}$ etc).
 
@@
 %$t\ t$				\hfill application				\alt
-$\filter(\strin[r],t)$ 		\hfill filter substrings			\alt
+$\filter{\strin[r]}{t}$ 		\hfill filter substrings			\alt
 $[\strin[r]](t)$ \hfill safe conversion \alt
-$\dconvert[\strin[r]](t)$ \hfill unsafe conversion
+$\dconvert{\strin[r]}{t}$ \hfill unsafe conversion
 
@@
 functions. In this treatment, we include concatenation and filtering.
-The $\filter$ function removes all instances of a regular expression in a string,
+The ${\tt filter}$ function removes all instances of a regular expression in a string,
 while concatenation ($+$) concatenates two strings.
@@
 \inferline{T-Convert}
-  {\Gamma \vdash [\strin[r]](e):\strin[r]}
-  {\Gamma \vdash e : \strin[r'] \\ \lang{r'} \subseteq \lang{r}}
+{\Gamma \vdash [\strin[r']](e):\strin[r']}
+{\Gamma \vdash e : \strin[r] \\ \lang{r'} \subseteq \lang{r}}
 
 \inferline{T-Dconvert (optional)}
-{\Gamma \vdash \dconvert[\strin[r]](e):\strin[r]}
-{\Gamma \vdash e : \strin[r']}
+{\Gamma \vdash \dconvert{\strin[r']}{e}:\strin[r']}
+{\Gamma \vdash e : \strin[r]}
 
 \inferline{T-Filter}
-  {\Gamma \vdash \filter(\strin[r'], t) : \strin[(r \backslash r') + \emptyset]}
+{\Gamma \vdash \filter{\strin[r']}{t} : \strin[(r \backslash r') + \emptyset]}
 {\Gamma \vdash t:\strin[r]}
@@
 \inferline{E-filterval}
-{\filter(\strin[r], e) : \strin[r'] \reduces {\tt rl\_filter}(r,e)}
+{\filter{\strin[r']}{e} : \strin[r\backslash r' + \emptyset] \reduces {\tt rl\_filter}(r,e)}
 { e \val }
@@
 \inferline{E-filter}
-{\filter(\strin[r], e) : \strin[r'] \reduces  \filter(\strin[r], e' )}     
-  { e : T \reduces e' }
+{\filter{\strin[r']}{e} : \strin[r\backslash r' + \emptyset] \reduces  \filter{\strin[r]}{ e' }}     
+{ e : \strin[r] \reduces e' }
+
+\inferline{E-convertval}
+{ [\strin[r']](e) : \strin[r'] \reduces [\strin[r']](e') }
+{ e : \strin[r] \reduces e' }
+
+\inferline{E-convert}
+{ [\strin[r']](e) : \strin[r'] \ireduces \istrf{e} }
+{ e \val }
 
 \inferline{E-dconvertval (optional) \footnote{See definition ???} }
-{\dconvert[\strin[r]](e) : \strin[r'] \reduces {\dconvert[\strin[r]](e'})}
-{e:T \reduces e'}
+{\dconvert{\strin[r']}{e} : \strin[r'] \reduces {\dconvert{\strin[r']}{e'}}}
+{e:\strin[r] \reduces e'}
 
 \inferline{E-dconvert (optional) \footnote{See definition ???} }
-{\dconvert[\strin[r]](e) : \strin[r'] \ireduces {\tt ifilter}(r, \istrf{e})}
+{\dconvert{\strin[r']}{e} : \strin[r'] \ireduces \ifilter{r'}{\istrf{e}}}
 {e \val }
@@
 
-\begin{defn}[Type translation]
+\begin{defn}[Type Translation Function $h$]
   The type translation function $h : Type \rightarrow IType$ is defined as follows:
@@
 
+Before presenting progress and preservation theorems, we establish some definitions
+and lemmas necessary for the proofs.
+
+\begin{defn}[Types of internal values]
+The types of internal values are defined as follows:
+\begin{itemize}
+  \item If $e=\istrf{s}$ then $e:\istr$.
+  \item If $e=\ifilter(r, \istrf{s})$ then $e:\istr$.
+  \item If $e=\istrf{s_1}+\istrf{s_2}$ then $e:\istr$.
+\end{itemize}
+\end{defn}
+
+Definition 2 is an additional assumption about
+the internal language. To summarize, we assume an internal language containing 
+strings together with operations for string concatentation and filtering. We
+expect closure over strings for both operations.
+Recall that $\ifilter$ is only needed for dynamic casts, which may be removed without
+descreasing the expressivity or even usability of the language.
+
+%\begin{lem}[Canonical Forms for $\lcs$]
+%We identify only canonical forms necessary for the type safety proof.
+%\begin{itemize}
+%  \item If $e \val$ and $e:\str$ then $e = \strf{s}$.
+%\end{itemize}
+%\end{lem}
+
 \begin{thm}[Preservation]
@@
 The proof is a straightforward induction on the derivation of the combined evaluation relation.
-%\begin{itemize}[label=$ $,itemsep=1ex]
-%  \item \textbf{E-Ival, E-Ifilterval}. Both cases hold vacuously since the terms at hand are not $\lcs$ terms.
-%  \item \textbf{E-String}. Since $e:\str$, it suffices to show that $h(\str) = \istr$ which holds by definition.
-%  \item \textbf{E-Sring\_in}. Since $e:\strin[r]$, it suffices to show that $h(\strin[r]) = \istr$ which holds for any $r$ by definition.
-%  \item \textbf{E-filter}. Note that ${\tt filter}(\strin[r], e) : \strin[r']$ for some $r'$. 
-%    It suffices to show that rl\_filter(r,e)
-%    $h(\strin[r]) = \istr$ which holds for any $r$ by definition.
-%  \item \textbf{E-Dconvert}. Note that $\dconvert[r](e) : \strin[r]$, so it suffices to show that $h(\strin[r]) = \istr$ which holds by defintiion.
-%\end{itemize}
+\begin{itemize}[label=$ $,itemsep=1ex]
+  \item \textbf{E-Ival, E-Ifilterval}. Both cases hold since the terms at hand are not $\lcs$ terms.
+  \item \textbf{E-Stringval, E-strval}. Both cases hold since no reduction is possible. 
+  \item \textbf{E-String}. By the definition of typing for internal terms, $\istrf{e} : \istr$. It suffices to show that $h(\str) = \istr$, which follows from the definition of $h$.
+  \item \textbf{E-String\_in}.
+By the definition of typing for internal terms, $\istrf{e} : \istr$.
+It suffices to show that $h(\strin[r]) = \istr$, which follows from the definition of $h$ for arbitrary $r$.
+  \item \textbf{E-concat}.
+By the definition of typing for internal terms, $\istrf{e_1} + \istrf{e_2} : \istr$.
+So it suffices to show that $h(\strin[r_1+r_2]) = \istr$, which follows from the definition of $h$ for arbitrary $r_1$, $r_2$.
+  \item \textbf{E-Filterval}. We have that $\filter{\strin[r']}{e}:\strin[r\backslash r' + \emptyset]$. By inversion of T-Filter, $e:\strin[r]$.
+    By T-Equiv-string\_in (which is bidirectional), $e \in \lang{r}$.
+    By theorem 2, ${\tt rl\_filter}(r,e) \in \lang{r\backslash r' + \emptyset}$.
+    By T-Equiv-string\_in,  $e \in \lang{r}$ and ${\tt rl\_filter}(r,e) \in \lang{r\backslash r' + \emptyset}$ implies ${\tt rl\_filter}(r,e):\strin[r\backslash r' + \emptyset]$.
+  \item \textbf{E-Filter}. By inversion, $e : \strin[r'] \reduces e'$ so by the induction $e':\strin[r']$.
+    Therefore, ${\tt filter}(\strin[r], e') : \strin[r \backslash r' + \emptyset]$ by T-Filter.
+  \item \textbf{E-Convertval}. It suffices to show that $h(\strin[r]) = \istr$, which is true by definition.
+  \item \textbf{E-Convert}. By inversion and induction, $e' : \strin[r]$. We know that $[\strin[r']](e) : \strin[r']$, so by inversion of T-Convert $\lang{r'} \subseteq \lang{r}$.
+    It follows that $[\strin[r']](e') : \strin[r']$.
+  \item \textbf{E-DConvert}. By inversion, $e : \strin[r'] \reduces e'$. By the induction hypothesis, $e' : \strin[r']$; therefore, by T-Dconvert, $\dconvert[\strin[r]](e):\strin[r]$.
+  \item \textbf{E-DConvertval}.
+By the definition of typing for internal terms, $\ifilter(r,\istrf{e}) : \istr$.
+It suffices to show that $h(\strin[r_1+r_2]) = \istr$, which follows from the definition of $h$.
+\end{itemize}
 \end{proof}
@@
 \begin{thm}[Progress]
-  Let $T$ be a type in $\lcs$ and $\sigma$ a type in the internal language.
-  For all terms $e$, if $e:T$ then $e:T \reduces^* e' \ireduces^* i$ where $i {\tt ival}$.
+  If $e:T$ then $e:T \reduces^* e' \ireduces^* i$ where $i \ival$.
 \end{thm}
@@
 By induction on the derivation of $e:T$.
-%\begin{itemize}[label=$ $, itemsep=1ex]
-%\end{itemize}
+\begin{itemize}[label=$ $,itemsep=1ex]
+  \item \textbf{T-Equiv-Include}. Since $e \in \lang{r}$, $e = \strf{s}$ for some $s$; therefore, $e \val$ by E-Strinval.
+  \item \textbf{T-Minimum}. 
+\end{itemize}
 \end{proof}

### 2014-03-25 ace-oopsla14/string-appendix.tex
Hand-wavy progress proof. Moving on to cleanup and regex section for now.

@@
     By T-Equiv-string\_in (which is bidirectional), $e \in \lang{r}$.
-    By theorem 2, ${\tt rl\_filter}(r,e) \in \lang{r\backslash r' + \emptyset}$.
+    By the Regular Language Intersection Theorem, ${\tt rl\_filter}(r,e) \in \lang{r\backslash r' + \emptyset}$.
     By T-Equiv-string\_in,  $e \in \lang{r}$ and ${\tt rl\_filter}(r,e) \in \lang{r\backslash r' + \emptyset}$ implies ${\tt rl\_filter}(r,e):\strin[r\backslash r' + \emptyset]$.
@@
   \item \textbf{T-Equiv-Include}. Since $e \in \lang{r}$, $e = \strf{s}$ for some $s$; therefore, $e \val$ by E-Strinval. 
-  \item \textbf{T-Minimum}. 
+    We have that $e \val$ and $e : \strin[r]$, so $e \ireduces \istrf{e}$ by E-string\_in.
+  \item \textbf{T-Minimum}. Result follows immediately by induction and E-string\_in.
+  \item \textbf{T-Concat}. Result follows immediately by induction and E-concat.
+  \item \textbf{T-Convert}. Result follows immeidately by induction, E-Convert and E-Convertval.
+  \item \textbf{T-Dconvert}. Result follows immediately by induction, E-Dconvertval and E-Dconvert.
+  \item \textbf{T-Filter}. Result follows immediately by induction and E-Filter.
 \end{itemize}

### 2014-03-25 ace-oopsla14/ace-oopsla14.tex

@@
 \usepackage{verbatimbox}
-
+\input{../att-icfp14/macros-atlam}
 \usepackage{alltt}
@@
 
-\title{\Ace: Growing A Statically-Typed Language Inside a Python}
+\title{\Ace: An Extensible Static Type System Inside a Python}
 
@@
 Python serves as Ace's type-level language.
-Rather than building in a particular type system, Ace is organized around an extensible bidirectional type system that inverts control over typechecking and translation to type-level functions according to a protocol that cannot lead to ambiguities and type safety issues when type systems are combined.
-We give a simplified calculus that describes the type theoretic foundations of this mechanism and show that, as implemented in Ace, it is flexible enough to admit practical library-based implementations of types drawn from functional, object-oriented, parallel and domain-specific languages.
-We also show how to construct safe, natural and extensible foreign function interfaces using a novel form of ``staged'' type propagation from Python into Ace.
+Rather than building in a particular set of type constructors, Ace introduces a novel extension mechanism, \emph{active typechecking and translation}, organized around a bidirectional type system that inverts control over typechecking and translation to user-defined type constructors according to a protocol that cannot cause ambiguities and type safety issues at link-time.
+In addition to describing a full-scale language design, we give a simplified calculus that describes the foundations of this mechanism and show that, as implemented in Ace, it is flexible enough to admit practical library-based implementations of types drawn from functional, object-oriented, parallel and domain-specific languages.
+We also show how to construct safe, natural and extensible foreign function interfaces using a novel form of ``staged'' type propagation from Python into Ace and then through to OpenCL.
 % can be used to enable type system extensions and demonstrates how type safety can be maintained by lifting a technique from the typed compilation literature into the language.%To show that the method scales, we give a full implementation of the OpenCL programming language as a library.
@@
 
-%\category{D.3.2}{Programming Languages}{Language Classifications}[Extensible Languages]
+\category{D.3.2}{Programming Languages}{Language Classifications}[Extensible Languages]
 %\category{D.3.4}{Programming Languages}{Processors}[Compilers]
@@
 
-In this paper, we describe an extensible statically-typed language, Ace, that is implemented entirely as a library within Python, avoiding the ``chicken-and-egg problem'' and occupying what we intend to convey by example as a ``sweet spot'' in the design space (contribution 1, Sec.
+In this paper, we describe an extensible statically-typed language, Ace, that is implemented entirely as a library within Python and flexibly repurposes its syntax to enable, for example, type annotations.
+This avoids the ``chicken-and-egg problem'' and occupies what we consider a ``sweet spot'' in a design space that includes the extrinsic criteria described above.
+We justify this by showing a variety of non-trivial type system fragments that can be embedded as composable libraries within Ace (contribution 1, Sec.
 \ref{usage}).
 
-Ace, together with a core lambda calculus, serves also as a vehicle for introducing a novel extension mechanism we call \emph{active typechecking and translation}.
-We sidestep the expression problem entirely by leaving the forms of expressions fixed, sharing them with Python.
+Ace, together with a core lambda calculus, serves also as a vehicle for a more general contribution: a novel  underlying extension mechanism, which we call \emph{active typechecking and translation}.
+As suggested by the design of Ace, it sidestep the expression problem by leaving the forms of expressions fixed.
 Instead, abstraction providers extend the language by introducing new type constructors and operators using type-level functions.
-We show how, by repurposing the conventional syntactic forms found in languages like Python, structuring the language as a bidirectional type system and integrating techniques from the typed compilation literature directly into the language, we can maintain expressiveness while  guaranteeing that link-time ambiguities cannot occur and type safety is maintained (contribution 2, Secs.
+We show how, by  structuring the language as a bidirectional type system and integrating techniques from the typed compilation literature directly into the language, we can maintain expressiveness while  guaranteeing that link-time ambiguities cannot occur and type safety is maintained (contribution 2, Secs.
 \ref{att} and \ref{theory}).
@@
 Python itself, as well as the others just described) can also be compiled and invoked interactively from Python.
-Ace supports a form of \emph{ad hoc} polymorphism (based on singleton types) whereby the ``dynamic type'' of a Python value can determine a static type ``just-in-time'' when Python code calls a generic Ace function.
-This represents a form of implicit \emph{staging} and we show how this can significantly streamline interactive workflows (like those in scientific computing) that alternate between Python for orchestration and exploratory tasks (e.g.
-plotting) and a statically-typed target  language via a foreign-function interface (FFI) for performance- and correctness-critical portions.
-We discuss a complete implementation of the OpenCL type system (a language for low-level data parallel programming on GPUs), show several higher-level extensions to it, and discuss deep bindings between Ace, the OpenCL run-time system and Python's standard numerics package, \verb|numpy|, all implemented within a library (contribution 3, Sec.
+Ace supports a form of staged \emph{ad hoc} polymorphism (based on singleton types) wherein the ``dynamic type'' of a Python value can determine a static type and trigger active typechecking and translation ``just-in-time'' when Python code calls a generic Ace function.
+We show how this can significantly streamline interactive workflows (e.g.
+in scientific computing) that use Python for orchestration and exploratory tasks (e.g.
+plotting) but call into a statically-typed  language via a foreign-function interface (FFI) for performance- and safety-critical portions.
+We provide a complete implementation of the OpenCL type system (a language for low-level data parallel programming on GPUs), show several higher-level extensions to it, and discuss bindings that bridge between the type systems of Ace, OpenCL and Python's standard numerics package, \verb|numpy|, all as a library (contribution 3, Sec.
 \ref{targets}).%We will discuss in detail as our primary realistic case study.
@@
 In Sec.
-\ref{related}, we compare Ace to related work on language extensibility, term rewriting and staged compilation.
+\ref{related}, we place Ace in the context of related work on language extensibility, term rewriting, staged  compilation, bidirectional typechecking, type-level computation and typed compilation.
 We conclude in Sec.
-\ref{discussion} by highlighting the key features needed by a host language to support active typechecking and translation, and discussing limitations and potential future work.
-The appendix contains details omitted for concision as well as several more examples, including functional datatypes, flexible complex numbers and a previously little-explored type system for tracking string invariants using regular expressions statically (which we show being used in Listing \ref{example}, amongst several other examples).
+\ref{discussion} by summarizing the key features needed by a host language to support active typechecking and translation, and discussing present limitations and potential future work.
+The appendix has details omitted for concision as well as several more examples, including functional datatypes, flexible complex numbers and a previously little-explored type system for tracking string invariants using regular expressions statically (which we show being used in Listing \ref{example}, amongst  other examples).
 
@@
 %
-Listing \ref{example} is an example of an Ace \emph{compilation script}.
-As promised, compilation scripts are written directly in Python.
-Ace requires no modifications to the language or features specific to the primary implementation, CPython (so Ace supports alternative implementations like Jython and PyPy).
+Listing \ref{example} shows that the top-level of every Ace file is a \emph{compilation script} written directly in Python.
+Ace requires no modifications to the language (version 2.6+ or 3.0+) or features specific to the primary implementation, CPython (so Ace supports alternative implementations like Jython and PyPy).
 This choice pays immediate dividends on the first five lines: Ace's import mechanism is Python's import mechanism, so Python's build tools (e.g.
 \verb|pip|) and package repostories (e.g.
-\verb|PyPI|) are directly available for distributing Ace-based libraries.
-Note that all of the type systems that we discuss are implemented entirely as  libraries.
+\verb|PyPI|) are directly available for distributing Ace-based libraries, including those defining type systems.
+%Ace can be installed with a single command: \emph{omitted for review}.
 %Ace is agnostic about the particular design choices that we made in our examples.
@@
 \begin{enumerate}
-\item On line 7, we construct a functional record type with a single (immutable) field named \verb|amount| with type \verb|decimal[2]|, which classifies decimal numbers with two decimal places.
-Here, \verb|record| and \verb|decimal| are \emph{indexed type constructors}.
-We will see how they are implemented in Sec.
-\ref{att}, but note that syntactically, \verb|record| borrows Python's syntax for array slices (e.g.
-\verb|a[min:max]|).
-Note that the field name could equivalently have been written \lstinline{'a' + 'mount'}, again emphasizing that types and indices are constructed programatically by a Python script.
-\item On lines 8-11, we construct another record type.
+\item On line 9, we construct a functional record type with a single (immutable) field named \verb|amount| with type \verb|decimal[2]|, which classifies decimal numbers with two decimal places.
+\verb|record| and \verb|decimal| are \emph{indexed type constructors}.
+We will provide more details in Sec.
+\ref{att}, but note that syntactically, \verb|record| overloads subscripting and borrows Python's syntax for array slices (e.g.
+\verb|a[min:max]|) to approximate conventional functional notation for type annotations.
+Note also that the field name could equivalently have been written \lstinline{'a' + 'mount'}, again emphasizing that types and indices are constructed programatically by a Python script.
+\item On lines 10-13, we construct another record type.
 The field \verb|name| has type \verb|string|, defined in \verb|examples.py|, while the fields \verb|account_num| and \verb|routing_num| have more interesting types, classifying strings guaranteed statically to be in a regular language specified statically by a  regular expression pattern (written, to avoid needing to escape backslashes, using Python's \emph{raw string literals}).
@@
 %Prototypic inheritance  is a feature of some dynamically-typed languages (most notably, Self \cite{Ungar:Smith:oopsla:1987} and Javascript). 
-On line 12, we construct a simple \emph{prototypic object type} \cite{Lie86}.
+On line 14, we construct a simple \emph{prototypic object type} \cite{Lie86}.
 The type \verb|Transfer| classifies terms consisting of a \todo{what is the right term for this?}\emph{fore} of type \verb|A| and a \emph{prototype} of type \verb|C|.
@@
 \end{enumerate}
-\subsection{Functions}
-
-Functions implementing run-time behavior (\emph{typed functions}) are distinguished by the presence of a decorator specifying their  \emph{base semantics} (here \verb|py| from \verb|examples.py|; we will discuss this further below).
+\subsection{Typed Functions}
+
+\emph{Typed functions} implement the run-time behavior and are distinguished by the presence of a decorator specifying their  \emph{base semantics} (here \verb|py| from \verb|examples.py|; Sec.
+\ref{att}).
 These functions are, however, still written using Python's syntax, a choice that is again valuable for extrinsic reasons: users of Ace can use a variety of  tools designed to work with Python source code  without  modification, including code highlighters (like the one used in generating this paper), editor plugins, style checkers and documentation generators.
-We will see how, with a bit of cleverness, Python's syntax can be repurposed within these functions to support a variety of static and dynamic semantics that differ from Python's.
-Ace  leverages the \verb|inspect| and \verb|ast| modules in the Python standard library to extract abstract syntax trees for typed functions \cite{python}.
-
-On lines 14-18, we see the function \verb|log_transfer|.
-As in Python, it begins with a documentation string  that informally specifies the behavior of the function.
+We will see several examples of how, with a bit of cleverness, Python's syntax can be repurposed within these functions to support a variety of static and dynamic semantics that differ from Python's.
+Ace  leverages the \verb|inspect| and \verb|ast| modules in the Python standard library to extract abstract syntax trees for functions annotated in this way \cite{python}.
+
+On lines 16-21, we see the function \verb|log_transfer|.
+We follow Python conventions by starting with a documentation string  that informally specifies the behavior of the function.
 Before moving into the body, however, we  also write a \emph{type signature} (line 17) stating that the type of \verb|t| is \verb|Transfer|, the prototypic object type described above.
@@
 
-Line 17 borrows Python's syntax for dictionary literals to approximate  conventional notation for type annotations.
+Line 19 repurposes Python's syntax for dictionary literals to approximate  conventional notation for type annotations.
 %\footnote{In the rare situation where there are no type annotations and the first statement is a dictionary literal, users can use Python's \texttt{pass} keyword.} % (and following conventions for documentation strings, which also describe ``types'' at the end).
@@
 \subsection{Bidirectional Typechecking of Introductory Forms}
-On lines 21-30, we define the function \verb|__toplevel__| (the base will consider this a special name for the purposes of compilation, discussed in the Sec.
+On lines 23-32, we define the function \verb|__toplevel__| (the base will consider this a special name for the purposes of compilation, discussed in Sec.
 \ref{compilation}).
-After printing a run-time greeting, we construct a \emph{value} of type \verb|C| on lines 24-26.
+After printing a run-time greeting, we introduce a value of type \verb|C| on lines 26-28.
 Recall that \verb|C| is a record type, so we provide the names of the fields (here, without quotes because we are no longer in the type-level language) and values of the appropriate type using the syntactic form normally used for dictionary literals.
@@
 This again repurposes existing  syntax: here function application.
-When the ``function'' is of literal form (dictionaries, tuples, lists, strings, numbers, booleans and  \verb|None|) and the argument is a type, we consider it instead an \emph{ascribed introductory form} of that type.
-We see another such form used with Python's tuple syntax on lines 27-28 to introduce terms of type \verb|Transfer| by providing the fore and prototype.
+When the ``function'' is of literal form (dictionaries, tuples, lists, strings, numbers, booleans and  \verb|None|) and the argument is a type, we will treat it instead an \emph{ascribed introductory form} of that type.
+We see another such form used with Python's tuple syntax on lines 29-30 to introduce terms of type \verb|Transfer| by providing the fore and prototype.
 
 %\subsection{Bidirectional Typechecking}
-The string, number and dictionary literal forms inside the outermost forms on lines 24-28 are also  introductory, but do not have an ascription.
-This is because the definition of the type \verb|C| completely determines which types they will need to have.
+The string, number and dictionary literal forms inside the outermost forms on lines 26-30 are also  introductory, but do not have an ascription.
+This is because the outermost ascriptions completely determine which types they will need to have.
 As we will discuss in more detail shortly, Ace is built around a \emph{bidirectional type system} that distinguishes locations where an expression must \emph{synthesize} a type (e.g.
@@
 as an argument to a function with known type) \cite{Lovas08abidirectional}.
-Literal forms are only analytic.
-
-If a literal is used in a synthetic location, the base can, however, specify a ``default ascription''.
+Unascribed literal forms must ultimately be analyzed against a type.
+
+If an unascribed literal is used in a synthetic location, the base can, however, specify a ``default ascription''.
 For example, if we had not specified the literal ascription on line 26, the base would cause the dictionary literal to be analyzed against the type \verb|dyn|, covering dynamically classified Python values.
-This would still lead to a type error because the variables used as keys are not bound in the top-level context (emphasizing that ``dynamically-typed languages'' can still have a static semantics, even if it only involves checking that top-level variables are bound).% When considered as a record, the keys were considered in a different class by the implementation (labels, rather than variables).
-
-An ascription can also evaluate statically to a {type constructor}.
+This would still lead to a type error because keys are then treated as expressions, as in Python, rather than field names, and the identifiers shown have not been bound in the top-level context (emphasizing that ``dynamically-typed languages'' can still have a static semantics, even if it only involves checking that top-level variables are bound).% When considered as a record, the keys were considered in a different class by the implementation (labels, rather than variables).
+
+An ascription can also be a {type constructor} instead of a type.
 For example, we might write \verb|[1, 2] (matrix)| instead of \verb|[1, 2] (matrix[i32])| when using a base for which the default ascription for number literals is \verb|i32|.
-Because the arguments are synthetic, the type constructor can determine an appropriate index based on the types of the subexpressions.
-If we wanted a matrix of \verb|f32|s, then we would have to write either \verb|[1, 2] (matrix[f32])| or \verb|[1(f32), 2(f32)] (matrix)|, or use a base with a different default ascription.
-We will see how bidirectional typechecking of introductory forms  works in greater technical detail in Secs.
-\ref{att} and \ref{theory}, but note here that it is one of the key mechanisms for expressiveness given our constraint that we cannot add entirely new forms to the language (and indeed, doing so would begin to be quite crowded even if we could add distinct forms for matrices, vectors, lists, tuples, etc.).
+Because the arguments synthesize the appropriate type, the type constructor can synthesize an appropriate index based on the types of the subexpressions.
+If we wanted a matrix of \verb|dyn|s, then we would have to write either \verb|[1, 2] (matrix[f32])|, ascribe each inner literal so that it synthesizes the \verb|f32| type, \verb|[1(f32), 2(f32)] (matrix)|, or construct a base with a different default ascription.
+
+We will see how bidirectional typechecking  works in greater technical detail in Secs.
+\ref{att} and \ref{theory}, but note here that having the semantics of a form change depending on the type it is being analyzed against is a key to achieving expressiveness given our constraint that we cannot add  new forms to the language, for both extrinsic and intrinsic reasons.
 
@@
 
-To typecheck, translate and execute the code in Listing \ref{example}, we have two choices: do so externally at the shell, or perform compilation interactively from within Python.
-We will begin with the former, then return to the latter in Sec.
+To typecheck, translate and execute the code in Listing \ref{example}, we have two choices: do so externally at the shell, or perform compilation interactively (or implicitly) from within Python.
+We will begin with the former, then return to interactive and implicit compilation in Sec.
 \ref{targets}.% with our OpenCL examples.
@@
 \begin{enumerate}
-\item Evaluate the provided compilation script (\verb|listing1.py|)
+\item It evaluates the compilation script to completion.
 \item For any top-level bindings that are Ace functions in the final environment (instances of \verb|ace.TypedFn|, as we will discuss), it initiates active type-checking and translation (Sec.
@@
 \ref{targets}.
-Here our target is  the default target associated with the base \verb|examples.py.py|, which emits Python 2.6 files.
+Here our target is  the default target associated with the base \verb|py|, which emits Python 2.6 files.
 If a type error is discovered, no file is emitted and the error is displayed on the console.
@@
 \item The base recognized the function name \verb|__toplevel__| as special, placing the translation of its body  at the top level of the file.
-Ace does not have any special function names itself; this is a feature of the chosen base.
+Ace does not have any special function names itself; this is a feature of the user-defined base.
 \item Records with two or more fields translated to tuples of their values (e.g.
@@
 \item Terms of type \verb|string_in[r"..."]| translated to strings.
-Checks for membership in the specified regular language were performed entirely statically by the type system.\todo{Nathan?}\footnote{Note that there are situations (e.g.
+Checks for membership in the specified regular language were performed entirely statically by the type system.\footnote{Note that there are situations (e.g.
 if a string is read in from the console) that would necessitate an initial run-time check, but no further checks in downstream functions would be necessary.
@@
 \end{codelisting}Listing \ref{oops} shows an example of code containing several type errors.
-Indeed, lines 9-13 each contain a type error.
-If analagous code were written in Python itself, these could only be found if the code was executed on the first day of the month (and, depending on the implementation, not all of the issues would  immediately result in run-time exceptions, possibly leading to quite subtle problems).
+Indeed, lines 8-12 each contain a type error.
+If analagous code were written in Python itself, these could only be found if the code was executed on the first day of the month (and, depending on the implementation, not all of the issues would  immediately result in run-time exceptions, possibly leading to quite subtle problems; for example, the unexpected use of \verb|None|\footnote{The notation \texttt{+T} constructs a \texttt{None}-able option type; see appendix.}).
 Static type checking allows us to find these errors during compilation.
@@
 The compilation script completes (so that functions can refer to each other mutually recursively), then the typed functions in the top-level environment are typechecked.
-The typechecker raises an exception statically at the first error, and \verb|acec| prints it as shown.
+The typechecker raises an exception statically at the first error, and \verb|acec| prints it to the console as shown.
 
@@
 
-In compilers and other tools for monolithic languages, there is typically a \emph{syntax-directed} protocol governing type checking, translation, pretty-printing and so on.
-For example, one might declare functional datatypes that capture all types and term forms, and then proceed by  exhaustive case analysis over terms to implement this functionality.
-The external visitor pattern for class-based object systems captures essentially the same protocol.
+%With monolithic languages, enabling a variety of tools is relatively straightforward.
+For example, one might declare functional datatypes with cases for every type constructor and term form, and then proceed by  exhaustive case analysis over term forms to implement this sort of functionality.
+The external visitor pattern for class-based object systems operates in essentially the same way.
 In both cases, the problem is that the types and terms are defined in one place, so  providers cannot extend the language modularly.
 
-In extensible languages the focus is typically on enabling the addition of both new types and term forms.
-Adding new term forms requires then adding cases for all possible functions over terms (e.g.
-typing, translation,  code highlighting, style checking and so on).
-This is difficult.
-Indeed, this is the canonical example of the widely-studied \emph{expression problem} \cite{wadler1998expression}.
-Although there are some approaches that work for certain purposes, retaining the ability to reason compositionally about metatheoretic issues (e.g.
-type safety) and avoiding ambiguities when extensions are combined can become difficult (see Sec.
+Enabling the addition of new forms to a language in an open tool ecosystem is difficult.
+Indeed, it is the canonical example of the long-studied \emph{expression problem} \cite{wadler1998expression}.
+Although a number of approaches have been developed, we argue that many of them are overly permissive, making it difficult to reason compositionally about metatheoretic issues (e.g.
+type safety) and avoid ambiguities when extensions are combined (see Sec.
 \ref{related}).
+%They are also often quite complex.
 %Requiring that abstraction clients have the expertise to resolve these ambiguities manually is undesirable, and heading them off results in an unbounded, combinatoric explosion of situations to consider.
-
 Ace's term forms, as we have seen, are fixed by Python's grammar, so Ace sidesteps the expression problem almost entirely (and thus inherits broad tool support, as described earlier).
 The key insight is that leaving the forms fixed does not mean the semantics must also be fixed.
-Instead of taking a syntax-directed view of extensibility, we take a type-directed view: users cannot add new forms but they can add new types, which we will allow to affect the semantics of or reinterpret existing forms (e.g.
+Instead of taking a syntax-directed view of extensibility, we take a type-directed view: users cannot add new forms but they can add new type constructors, which we empower with  more control over the semantics of existing forms (e.g.
 literals, as discussed above, but also nearly every other form, as we will now discuss) in a controlled manner.
@@
 \ref{theory}), users can also repurpose existing forms flexibly within a delimited scope.
-More specifically, the key features that distinguish active typechecking and translation are: \begin{itemize}
-\item a \emph{dispatch protocol} that delegates responsibility for typechecking and translation to:
+The key features that characterize active typechecking and translation are: \begin{itemize}
+\item a \emph{dispatch protocol} that delegates responsibility for typechecking and translating a term to:
   \begin{itemize}
-  \item a type or type constructor extracted from a subterm of a compound expression
-  \item a per-function base semantics (or simply \emph{base}) for base forms for which this is not possible  (statements, variables and unascribed literals)
+  \item a type or type constructor extracted from a subterm
+  \item a per-function base semantics (or simply \emph{base}) for forms for which this is not possible  (variables, literals without ascriptions and most statements)
   \end{itemize}
-\item a mechanism for allowing the user to define new types, type constructors and bases from within the language (in particular, from within the type-level language) rather than fixing them ahead of time
+\item a mechanism for allowing the user to define new types, type constructors and bases from within the language (in particular, \emph{from within the type-level language}) rather than fixing them ahead of time
 \end{itemize}
@@
   x = x + x
-  return {amount: x} (A)
+  return {
+    y: {amount: x} (A),
+    remark: "Creating an anonymous record"} (record)
 \end{lstlisting}
@@
 \subsection{Base Decorators}
-Function decorators in Python (line 2) are syntactic sugar for {calling} the decorator (\verb|py|) with the decorated function (lines 3-5) as an argument, then replacing the binding (\verb|example|)  with the return value of this call.
-We could equivalently have omitted line 3 and inserted the top-level statement \verb|example = py(example)| on the line after the function definition.
+Decorators in Python (line 2) are syntactic sugar: we could equivalently have omitted line 3 and inserted the top-level statement \verb|example = py(example)| on the line after the function definition.
 Note, however, that \verb|py| is not a function but an instance of a class, \verb|examples.py.PyBase|, that inherits from \verb|ace.Base|.
 It can be used as a decorator because \verb|ace.Base| overloads the call operator, shown in Listing \ref{base}.
-The \verb|_aceify| function (not shown) \todo{show in appendix?}operates as follows:
+The \verb|_process| function (not shown) \todo{show in appendix?}operates as follows:
 \begin{enumerate}
 \item The Python standard library is used to extract an abstract syntax tree (\verb|AST|) from the function.
-\item The closure of the function, together with the globals in the module it is defined in, are extracted to form its static environment.
+\item The closure of the function, together with the globals in the module it is defined in, are extracted to reify its static environment.
 \item The argument annotations are processed.
-If provided in the body, as in our example, it checks that the argument names are spelled correctly and evaluates the type-level expressions (e.g.
+If provided in the body, as in our example, \verb|_process| checks that the argument names are spelled correctly and evaluates the type-level expressions (e.g.
 \verb|A| above) in the static environment to produce a mapping of argument names to types, called the argument signature.
-If Python 3's argument annotations are used, this has  already been done for us.
+If Python 3's annotations were used, this is not necessary.
 %\item Whether the function is generic or not is determined.  We defer discussion of generic functions until Sec. \ref{targets}.
@@
 
-\subsection{Typechecking Typed Functions and Statements}
+\subsection{Checking Typed Functions}
 When the Ace compiler begins typechecking a \verb|TypedFn| (when asked to by \verb|acec|, or when it is  invoked interactively as we will show in Sec. \ref{targets}), it proceeds as follows:
@@
 \item The base is asked to initialize the context.
-This can be seen on lines 5-7 of Listing \ref{pybase}, where the Python base adds an attribute tracking local bindings (initially only the arguments) and the return type to be synthesized (initially not known, indicated using \verb|None|).
-\item For each statement in the body (ignoring the documentation string and type annotation, if present), the compiler calls a method of the base, \verb|check_|$F$, where $F$ is the form of the statement (taken directly from Python's \verb|ast| package \cite{python}).
-These methods are responsible for checking that statement is well-typed (statements do not themselves have a type) and implementing any necessary effects on the context.
-
-The assignment statements on lines 5 and 6 of Listing \ref{att-example} trigger the method \verb|check_Assign| \todo{if time, add destructuring assignment to appendix}shown on lines 6-14 of Listing \ref{pybase} and the return statement on line 7 triggers the method \verb|check_Return| on lines 18-22.
-Both work similarly: if this is the first assignment to a particular identifier, or the first return statement seen, then the value must synthesize a type in the current context.
-Otherwise, it is analyzed against the previously synthesized type.
-\item The base synthesizes a type for the function as a whole via the \verb|syn_ty_FunctionDef_outer| method.
+This can be seen on lines 2-4 of Listing \ref{pybase}, where the \verb|PyBase| base adds an attribute tracking local bindings (initially only the arguments) and the return type, which will be synthesized but initially is not known.
+\todo{briefly mention return type annotations?}
+\item For each statement in the body (after the documentation string and type annotation, if present), the compiler calls the context's \verb|check| method, which in turn defers to a method of the base, \verb|check_|$F$, where $F$ is the form of the statement (taken directly from Python's \verb|ast| package \cite{python}).
+These methods are responsible for checking that the statement is well-typed (raising an \verb|ace.TypeError| if not) and having any necessary effect on the context.
+\item The base synthesizes a type for the function as a whole via the \verb|syn_FunctionDef_outer| method.
 Here, an \verb|arrow| type is synthesized (lines 26-29) based on the argument signature and the synthesized return type.
@@
 
-\subsection{Bidirectionally Typechecking Expressions}
+Because bases are user-defined objects, this represents the first instance of the compiler delegating control to a library.
+\subsection{Bidirectional Typechecking}
 \begin{codelisting}
@@
   def __call__(self, f):
-    (ast, static_env, arg_sig) = _aceify(f)
+    (ast, static_env, arg_sig) = _process(f)
     return TypedFn(self, ast, static_env, arg_sig)
@@
   def check_Assign(self, ctx, s):
-    if len(s.targets) > 1:
-      raise ace.TypeError("...not supported...", s)
+    ace.util.unsupported_if(len(s.targets) != 1
+      or not isinstance(s.targets[0], ast.Name))
     x, e = s.targets[0].id, s.value # see ast module
     if x in ctx.locals:
-	  ctx.ana_ty(e, ctx.locals[x])
+      ctx.ana(e, ctx.locals[x])
     else:
-	  ty = ctx.syn_ty(e)
+      ty = ctx.syn(e)
       ctx.locals[x] = ty
@@
     if ctx.return_t == None:
-      ctx.return_t = ctx.syn_ty(s.value)
+      ctx.return_t = ctx.syn(s.value)
     else:
-      ctx.ana_ty(s.value, ctx.return_t)
+      ctx.ana(s.value, ctx.return_t)
 
@@
 
-  def syn_ty_FunctionDef_outer(self, ctx, f):
+  def syn_FunctionDef_outer(self, ctx, f):
     if ctx.return_t == None:
@@
 
-  def syn_ty_Name(self, ctx, e):
+  def syn_Name(self, ctx, e):
     x = e.id
@@
     elif x in ctx.fn.static_env: # (Sec. 5)
-      return self.syn_ty_lifted(x)
+      return self.syn_lifted(x)
     else:
@@
   default_target = PyTarget()   
-  default_Dict_ascription = dyn
+  default_Dict_asc = dyn
   # ...
@@
 
-The novel protocol for typechecking expressions that we will now describe (and formalize in Sec.
+\begin{codelisting}
+\begin{lstlisting}
+class Context(object):
+  def __init__(self, fn):
+    self.fn = fn
+
+  def check(self, s):
+    if isinstance(s, ast.Assign):
+      self.fn.base.check_Assign(self, s)
+    # ... other statement forms similar
+
+  def syn(self, e):
+    if instanceof(e, ast.Name):
+      delegate = self.fn.base
+      ty = delegate.syn_Name(self, e)
+    elif instanceof(e, ast.Attribute):
+      delegate = self.syn(e.value)
+      ty = delegate.syn_Attribute(self, e)
+    # ... other compound forms similar (cf appendix)
+    elif isinstance(e, ast.Dict):
+      return self.ana(self, e, 
+        self.fn.base.default_Dict_asc)
+    # ... other unascribed literal forms similar
+    elif is_ascribed_Dict(e):
+      lit, delegate = get_lit(e)
+      if issubclass(delegate, Type): # tycon
+        ty = delegate.syn_Dict(self, lit)
+      else:
+        return self.ana_Dict(lit, delegate)
+    # ... other ascribed literal forms similar
+    e.delegate = delegate
+    e.ty = ty
+    return ty
+
+  def ana(self, e, ty):
+    if instanceof(e, ast.Dict):
+      ty.ana_Dict(self, e)
+    # ... other literal forms similar
+    else:
+      syn = self.syn(e)
+      if ty != syn:
+        raise TypeError("...syn/ana mismatch...", e)
+      return
+    e.delegate = e.ty = ty
+
+  def trans(self, target, e):
+    d = e.delegate
+    if instanceof(e, ast.Name): 
+      trans = d.trans_Name(self, target, e)
+    elif instanceof(e, ast.Attribute):
+      trans = d.trans_Attribute(self, target, e)
+	# ... other forms similar
+	e.trans = trans
+	return trans
+\end{lstlisting}
+\caption{The \texttt{ace.Context} class delegates typechecking and translation of expressions, depending on their form and sub-terms, to a base, a type or a type constructor.}
+\label{context}
+\end{codelisting}
+
+%Active types are the primary means for extending Ace with new abstractions.
+An active type, as mentioned previously, is an instance of a class implementing the \verb|ace.Type| interface.
+Listing \ref{record} shows an example of such a class: the \verb|clx.Cplx| class used in Listing \ref{compscript}, which implements the logic of complex numbers.
+The constructor takes as a parameter any numeric type in \verb|clx| (line \ref{cplx}.2).
+
+\begin{codelisting}
+\begin{lstlisting}
+@slices_to_sig
+class record(ace.Type):
+  def __init__(self, sig, anon=False):
+    self.sig, self.anon = sig, anon
+
+  def __eq__(self, other): 
+    if self.anon or other.anon: 
+      return self is other
+    else: 
+      return (isinstance(other, record) 
+        and self.sig.equiv(other.sig))
+
+  def ana_Dict(self, ctx, e):
+    for f, v in zip(e.keys, e.values):
+      if f.id in self.sig:
+        ctx.ana(v, self.sig[f.id])
+      else:
+        raise ace.TypeError("...bad field...", f)
+    if len(self.sig) != len(e.keys):
+      raise ace.TypeError("...missing field...", e)
+
+  @classmethod # (when ascription is a type constructor)
+  def syn_Dict(cls, ctx, e):
+    sig = Sig()
+    for f, v in zip(e.keys, e.values):
+      sig[f] = ctx.syn_ty(v)
+    return cls(sig, anon=True)
+
+  def trans_Dict(self, ctx, target, e):
+    if len(self.sig) == 1:
+      return ctx.trans(e.values[0])
+    else:
+      value_dict = dict(zip(e.keys, e.values))
+	  return target.Tuple(
+	    ctx.trans(target, value_dict[field])
+	    for field, ty in self.sig)
+
+  def syn_Attribute(self, ctx, e): 
+    if e.attr in self.sig: 
+      return self.sig[e.attr]
+    else:
+      raise ace.TypeError("...field not found...", e)
+
+  def trans_Attribute(self, ctx, target, e): 
+    if len(self.sig) == 1:
+      return ctx.trans(target, e)
+    else:
+      idx = idx_of(self.sig, e.attr)
+      return target.Subscript(
+        ctx.trans(target, e.value), target.Num(idx))
+\end{lstlisting}
+%
+%  def trans_type(self, target):
+%    if target.supports(examples.fp.EagerNAryProdLR):
+%      return target.EagerNAryProdLR()
+%    else if isinstance(target, examples.py.Dyn):
+%      return target.Dyn()
+\caption{The \texttt{examples.fp.record} type constructor.}
+\label{record}
+\end{codelisting}
+The assignment statements on lines \ref{att-example}.5\footnote{In this section, we will need to refer to code in Listings \ref{att-example}, \ref{pybase}, \ref{context} and \ref{record}, so we adopt this syntactic convention to refer to line numbers for concision.} and \ref{att-example}.6 trigger the method \verb|check_Assign| \todo{if time, add destructuring assignment to appendix}shown on lines 6-14 of Listing \ref{pybase} and the return statement on line 7 triggers the method \verb|check_Return| on lines 18-22.
+Both work similarly: if this is the first assignment to a particular identifier, or the first return statement seen, then the value must synthesize a type in the current context.
+Otherwise, it is analyzed against the previously synthesized type in the current context.
+
+%The novel protocol for typechecking expressions that we will now describe (and formalize in Sec.
 \ref{theory}) is what enables us to extend the language with new semantics modularly while sidestepping the expression problem.
-Let us look more closely at how lines \ref{att-example}.5-\ref{att-example}.7\footnote{In this section, we will need to refer to code in Listings \ref{att-example}, \ref{pybase}, \ref{context} and \ref{record}, so we adopt this syntactic convention to refer to line numbers for concision.} are typechecked.
-
-As stated above, line \ref{att-example}.5 is an assignment statement, so the method \verb|check_Assign| on lines \ref{pybase}.6-\ref{pybase}.14 is called.
-Because the locals dictionary in \verb|ctx| does not yet know what type \verb|x| should be, the base asks to synthesize a type for the expression being bound, \verb|a.amount|, by calling the \verb|ctx.syn_ty| method.
-
-\paragraph{Active Type Synthesis} This method is defined on lines \ref{context}.5-\ref{context}.26.
-
-\subsection{Translation}
-The decorator, documentation string and argument annotations, if present, are stripped away (having already been processed).
-
-Once typechecking a method is complete, the compiler enters the translation phase, where terms in the target language are generated from Ace terms.
-Terms in the target language are generated by calling methods of the \emph{active target} governing the function being compiled.
-The translation phase operates similarly to typechecking, using the dispatch protocol to invoke methods named \verb|trans_|$X$.
-These methods have access to the context and node just as during typechecking, as well as the active target (named \verb|target| here).
+Let us look more closely at how lines \ref{att-example}.5-\ref{att-example}.7 are typechecked.
+
+The method \verb|check_Assign| on lines \ref{pybase}.6-\ref{pybase}.14 is called.
+Because the locals dictionary in \verb|ctx| does not yet know what type \verb|x| should be, the base asks to synthesize a type for the expression being bound, \verb|a.amount|, by calling \verb|ctx.syn|.
+
+%\paragraph{Active Type Synthesis} 
+This method is defined on lines \ref{context}.5-\ref{context}.26.
+The relevant case of the conditional is the one on line \ref{context}.9, because \verb|a.amount| is of the form \verb|Attribute| according to Python's grammar.
+Control over this term is given to the type recursively synthesized for its value, \verb|a|.
+We recurse back into \verb|syn|, now taking the first branch, corresponding to terms of the form \verb|ast.Name|.
+Control over names is given to the base, as seen on line \ref{context}.7, by calling the \verb|syn_Name| method.
+We can see its implementation for the base we are using on lines \ref{pybase}.39-\ref{pybase}.46.
+The identifier \verb|a| is an argument to the function, which the base included in the initial locals dictionary, so we hit a base case and the type \verb|A| is synthesized.
+
+We can now pop back up to line \ref{context}.11, which can now delegates control over \verb|a.amount| to the type \verb|A|.
+Recall that \verb|A| was constructed using the \verb|record| constructor in Listing \ref{example}.
+The definition of this type constructor is shown in Listing \ref{record}.
+In Ace, type constructors are classes inheriting from \verb|ace.Type| and types are instances of such classes.
+Constructor indices are given through the class constructor, called \verb|__init__| in Python.
+Here, \verb|record| requires a signature, which is a mapping of field names to types (an instance of \verb|examples.fp.Sig|, which performs well-formedness checking, not shown).
+The class decorator \verb|slices_to_sig| provides the convenient notation shown used in Listing \ref{example} (by adding a metaclass to override the class object's subscript operator, not shown).
+The \verb|syn_Attribute| method is shown on lines \ref{record}.38-\ref{record}.42.
+It simply looks in the record signature for the provided attribute, so \verb|decimal[2]| is synthesized.
+We can now go back up to the assignment statement that triggered this chain of calls and see on line \ref{pybase}.14 that a binding for \verb|x| is added to the locals dictionary and checking of this statement thus succeeds (no \verb|ace.TypeError| was raised).
+
+The next statement also assigns to \verb|x|, but this time, the analysis branch is taken using the \verb|ana| method of \verb|Context|, shown on lines \ref{context}.28-\ref{context}.37.
+In Ace, analysis is only used for literal forms.
+For any other form, the context simply asks for an equal type to be synthesized (we will discuss in Sec.
+\ref{discussion} future work on integrating subtyping, where this would be relaxed).
+For concision, we leave the details of synthesis for binary operators to the appendix, which also shows the definition of the \verb|decimal| type constructor.\todo{do this based on cplx; change intro}
+
+The final statement in \verb|example| is a return statement.
+Because it is the first return statement encountered, it too requires that the returned value synthesize a type.
+Here the returned value is an ascribed literal.
+We can see on lines \ref{context}.17-\ref{context}.22 how this is handled.
+Because the ascription is a type, \verb|A|, rather than a type constructor, the literal is analyzed against that type.
+Lines \ref{context}.29-\ref{context}.30 show that this involves simply calling the type's \verb|ana_Dict| method, shown on lines \ref{record}.7-\ref{record}.14.
+This simply ensures that all and only the fields in the signature were provided, and analyzes the values against the corresponding type from the signature.
+
+\subsection{Active Translation}
+Once typechecking is complete, the compiler enters the translation phase, generating terms in the target language specified by the base\todo{should we separate them?}.
+This phase follows essentially an identical protocol, using \verb|trans_|$X$ methods paired with each \verb|check_|/\verb|syn_|/\verb|ana_|$X$  method.
+The typechecking phase saves the entity that was delegated control as an attribute, \verb|delegate|, as well as the type of each AST node, \verb|ty|, so that these need not be determined again.
+The protocol for expressions can be seen on lines \ref{context}.39-\ref{context}.47.
+
+Translation methods have access to the context and node just as during typechecking, as well as an object, \verb|target|, that mediates code generation (named \verb|target| here).
+Because we are targeting Python directly, most of the \verb|trans| methods implement direct translations (factored out into a helper function).
+The main methods of interest are \verb|trans_FunctionDef_outer| on lines \ref{pybase}.33-\ref{pybase}-37 and the two translation methods in Listing \ref{record}, which implement the logic described in Sec.
+\ref{compilation}.
 
@@
 
-\subsection{Active Types}\label{atypes}
-
-\begin{codelisting}
-\begin{lstlisting}
-class Context(object):
-  def __init__(self, fn):
-    self.fn = fn
-
-  def syn_ty(self, e):
-    if instanceof(e, ast.Name):
-      owner = self.fn.base
-      ty = owner.syn_ty_Name(self, e)
-    elif instanceof(e, ast.Attribute):
-      owner = self.syn_ty(e.value)
-      ty = owner.syn_ty_Attribute(self, e)
-    # ... other compound forms similar (cf appendix)
-    elif isinstance(e, ast.Dict):
-      return self.ana_ty(self, e, 
-        self.fn.base.default_Dict_ty)
-    # ... other literal forms similar
-    elif is_ascribed_Num(e):
-      owner = get_ascription(e)
-      if is_tycon(owner):
-        ty = owner.syn_ty_Num(self, e)
-      else:
-        return owner.ana_ty_Num(self, e)
-    # ... other ascribed literal forms similar
-    e.owner = owner
-    e.ty = ty
-    return ty
-
-  def ana_ty(self, e, ty):
-    if instanceof(e, ast.Dict):
-      ty.ana_ty_Dict(self, e)
-    # ... other literal forms similar
-    else:
-      syn_ty = self.syn_ty(e)
-      if ty != syn_ty:
-        raise TypeError("...syn/ana mismatch...", e)
-      return
-    e.owner = e.ty = ty
-
-  def trans(self, target, e):
-    if instanceof(e, ast.Name): 
-      trans = e.owner.trans_Name(self, target, e)
-    elif instanceof(e, ast.Attribute):
-      trans = e.owner.trans_Attribute(self, target, e)
-	# ... other forms similar
-	e.trans = trans
-	return trans
-\end{lstlisting}
-\caption{The context dispatches active typechecking and translation of expressions to the base or an appropriate type or type constructor.}
-\label{context}
-\end{codelisting}
-
-%Active types are the primary means for extending Ace with new abstractions.
-An active type, as mentioned previously, is an instance of a class implementing the \verb|ace.Type| interface.
-Listing \ref{record} shows an example of such a class: the \verb|clx.Cplx| class used in Listing \ref{compscript}, which implements the logic of complex numbers.
-The constructor takes as a parameter any numeric type in \verb|clx| (line \ref{cplx}.2).
-
-\subsubsection{Dispatch Protocol}
-\begin{codelisting}
-\begin{lstlisting}
-@slices_to_sig
-class record(ace.Type):
-  def __init__(self, sig):
-    _check_sig(sig)
-    self.sig = sig
-
-  def ana_ty_Dict(self, ctx, e):
-    for f, v in zip(e.keys, e.values):
-      if f.id in self.sig:
-        ctx.ana_ty(v, self.sig[f.id])
-      else:
-        raise ace.TypeError("...bad field...", f)
-    if len(self.sig) != len(e.keys):
-      raise ace.TypeError("...missing field...", e)
-
-  def trans_Dict(self, ctx, target, e):
-    if len(self.sig) == 1: # assuming non-empty
-      return ctx.trans(e.values[0])
-    else:
-      value_dict = dict(zip(e.keys, e.values))
-	  return target.Tuple(
-	    ctx.trans(target, value_dict[field])
-	    for field, ty in self.sig)
-
-  def syn_ty_Attribute(self, ctx, e): 
-    if e.attr in self.sig: 
-      return self.sig[e.attr]
-    else:
-      raise ace.TypeError("...field not found...", e)
-
-  def trans_Attribute(self, ctx, target, e): 
-    idx = idx_of(self.sig, e.attr)
-    return target.Subscript(
-      ctx.trans(target, e.value), target.Num(idx))
-\end{lstlisting}
-%
-%  def trans_type(self, target):
-%    if target.supports(examples.fp.EagerNAryProdLR):
-%      return target.EagerNAryProdLR()
-%    else if isinstance(target, examples.py.Dyn):
-%      return target.Dyn()
-\caption{The \texttt{record} type constructor.}
-\label{record}
-\end{codelisting}
-\subsubsection{Typechecking}
-When typechecking a compound expression or statement, the Ace compiler temporarily hands control to the object selected by the dispatch protocol by calling the method \verb|type_|$X$, where $X$ is  the name of the syntactic form, taken from the Python grammar \cite{python} (appended with a suffix in some cases).
-
-%Note that in each example, subexpressions are not \emph{automatically} assigned types.
-Types must be requested from the context.
-This is useful because it allows types to reinterpret subexpressions of terms they control, to supporting more complex constructs (e.g.
-pattern matching on functional datatypes, see Sec.
-\ref{datatypes}).
-
-\subsubsection{Translation}
-
 \subsection{Compositional Reasoning}\label{safety}
@@
 \section{Theoretical Foundations}\label{theory}
+\begin{figure}
+\small
+\begin{flalign}
+\small
+ & \family{Tuple}{\klist{\kTypeBlur}}{\\
+ & \quad \tops{new}{\kunit}{\_}{a}{\tlam{\_}{\kunit}{\tlam{a}{\klist{\kDen}}{\\
+ & \quad\quad
+ 	\tfold{\tvar{a}}{\tden{{()}}{\ttype{Tuple}{\tnil{\kTypeBlur}}}}{d}{b}{r}{\\
+			& \quad\quad\quad \tfamcase{\ttypeof{\tvar{r}}}{Ntuple}{i}{\\
+			& \quad\quad\quad\quad \tfold{\tvar{b}}{\tden{\itransof{\tvar{d}}}{\ttype{Ntuple}{\tcons{\ttypeof{\tvar{d}}}{\tvar{i}}}}\\& \quad\quad\quad\quad}{\_}{\_}{\_}{\tden{\\& \quad\quad\quad\quad\quad {(\itransof{\tvar{d}},\itransof{\tvar{r}})}}{\ttype{Ntuple}{\tcons{{\ttypeof{\tvar{d}}}}{\tvar{i}}}}}
+			}{\terr}
+	}
+ }}};\\
+ & \quad \tops{prj}{\kint}{i}{a}{\lambda \tvar{i}:\kint.\lambda \tvar{a}:\klist{\kDen}. 
+ 	\tapp{\tvar{arity1}}{\tapp{\tvar{a}}{\tlam{d}{\kDen}{\\
+	& \quad\quad\tfamcase{\ttypeof{\tvar{d}}}{Ntuple}{nl}{\\
+ & \quad\quad \tfold{\tvar{nl}}{\terr}{t1}{j}{\_}{\\
+ & \quad\quad \tfold{\tvar{j}}{\tifeq{\tvar{i}}{1}{\dint}{\tden{\itransof{d}}{\tvar{t1}}}{\terr}\\&\quad\quad\quad}{\_}{\_}{\_}{\\
+ & \quad\quad\quad (\tlam{p}{\kpair{\kDen}{\dint}}{\tifeq{\tsnd{\tvar{p}}}{\tvar{i}}{\dint}{\tfst{\tvar{p}}}{\terr}})\\
+ & \quad\quad\quad (\tapp{\tvar{foldl}}{\tapp{\tvar{nl}}{\tapp{(\tden{\tvar{x}}{\tvar{nt}},0)}{
+ 	\tlam{r}{\kpair{\kDen}{\dint}}{\tlam{t}{\kTypeBlur}{\tlam{ts}{\klist{\kTypeBlur}}{\\
+	& \quad\quad\quad\quad \tifeq{\tvar{i}}{\tsnd{\tvar{r}}}{\dint}{\tvar{r}}{\tdencase{\tfst{\tvar{r}}}{rx}{\_}{\\
+	& \quad\quad\quad\quad\quad \tifeq{\tvar{i}}{\tsnd{\tvar{r}}+1}{\dint}{\\
+	& \quad\quad\quad\quad\quad\quad \tifeq{\tvar{ts}}{\tnil{\kTypeBlur}}{\klist{\kTypeBlur}}{(\tden{\tvar{rx}}{\tvar{t}},\tvar{i})\\
+	& \quad\quad\quad\quad\quad\quad}{(\tden{\titerm{\ifst{\iup{\tvar{rx}}}}}{\tvar{t}},\tvar{i})}\\	
+	& \quad\quad\quad\quad\quad}{(\tden{\titerm{\isnd{\iup{\tvar{rx}}}}}{\tvar{t}},\tsnd{\tvar{r}}+1)}}{\terr}}}}}
+ }}})
+ }}{\terr}}}}}
+ }\\ & 
+%& \quad \tops{pr}a}{b}{c}{d}
+ }{d}{i}{(\tlam{i}{\klist{\kTypeBlur}}{
+ 	\\& \quad \quad \tfold{\tvar{i}}{\titype{\dunit}}{s}{j}{r}{
+ 		\tfold{\tvar{j}}{\titype{\trepof{\tvar{s}}}}{\_}{\_}{\_}{\\
+		&\quad\quad\quad
+ 		\titype{\dpair{\trepof{\tvar{s}}}{\dup{\tvar{r}}}}
+ 		}
+ 	}
+ })}
+\end{flalign}
+\caption{ABC}
+\label{tuple}
+\end{figure}
+\begin{figure}[t]
+\small
+$\fbox{\inferrule{}{\progOKX{\progsort}}}$
+~~~$\tvarCtx ::= \emptyctx \pipe \tvarCtxX{t}{\kappa}$
+~~~$\fvalCtx ::= \emptyctx \pipe \fvalCtxX{\fvalDf}$
+%$\fCtx ::= \Sigma_0	 \pipe \fvalCtxX$
+\begin{mathpar}
+\inferrule[k-tycon]{
+	\fvar{Tycon} \notin \text{dom}(\fvalCtx)\\
+	\kEq{\kappaidx}\\
+	\tKind{\tvarCtx}{\fvalCtx}{\taurep}{\karrow{\kappaidx}{\kIType}}\\
+	\opType{\tvarCtx}{\fvalCtxX{\fvalDf}}{\theta}{\Theta}\\
+	\progOK{\tvarCtx}{\fvalCtxX{\fvalDf}}{\rho}
+}{
+	\progOKX{\pfam{\familyDf}{\rho}}
+}
+
+%\inferrule[def-kinding]{
+%	\tKindX{\tau}{\kappa}\\
+%	\progOK{\tvarCtxX{t}{\kappa}}{\fvalCtx}{\rho}
+%}{
+%	\progOKX{\pdef{t}{\kappa}{\tau}{\rho}}
+%}
+%
+\inferrule[k-e-prog]{
+	\exprOK{\tvarCtx}{\fvalCtx}{e}
+}{
+	\progOKX{e}
+}
+\end{mathpar}
+$\fbox{$\opType{\tvarCtx}{\fvalCtx}{\theta}{\Theta}$}$
+\begin{mathpar}
+\inferrule[k-opcon]{
+	\tKind{\tvarCtx}{\fvalCtx}{\taudef}{\karrow{\kappaidx}{\karrow{\klist{\kDen}}{(\ksum{\kDen}{\kunit})}}}
+}{
+	\opType{\tvarCtx}{\fvalCtx}{{\tops{op}{\kappaidx}{i}{a}{\taudef}}}{
+	\kOpS{op}{\kappaidx}}
+}
+
+\inferrule[k-opcons]{
+	\opType{\tvarCtx}{\fvalCtx}{\theta_1}{\Theta_1}\\
+	\opType{\tvarCtx}{\fvalCtx}{\theta_2}{\Theta_2}\\\\
+	\text{dom}(\theta_1) \cap \text{dom}(\theta_2) = \emptyset
+}{
+	\opType{\tvarCtx}{\fvalCtx}{\theta_1; \theta_2}{\Theta_1, \Theta_2}
+}
+\end{mathpar}
+$\fbox{$\exprOKX{e}$}$
+\begin{mathpar}
+\inferrule[k-e-var]{ }{
+	\exprOK{\tvarCtx}{\fvalCtx}{\evar{x}}
+}
+
+\inferrule[k-e-lam]{
+	\tKindX{\tau}{\kTypeBlur}\\
+	\exprOK{\tvarCtx}{\fvalCtx}{e}
+}{
+	\exprOKX{\elam{\evar{x}}{\tau}{e}}
+}
+
+\inferrule[k-e-op]{
+	\fval{Tycon}{-}{-}{\theta} \in \fvalCtx\\
+	\tops{op}{\kappaidx}{i}{a}{-} \in \theta\\
+	\tKindX{\tauidx}{\kappaidx}\\
+	\exprOKX{e_1}\\
+	\cdots\\
+	\exprOKX{e_n}
+}{
+	\exprOKX{\eop{Tycon}{op}{\tauidx}{\splat{e}{1}{n}}}
+}
+\end{mathpar}
+$\fbox{$\kEq{\kappa}$}$
+\begin{mathpar}
+\inferrule[t-eq]{ }{
+	\kEq{\kTypeBlur}
+}
+
+\inferrule[i-eq]{ }{
+	\kEq{\kint}
+}
+
+\inferrule[l-eq]{ }{
+	\kEq{\klabel}
+}
+
+\inferrule[list-eq]{
+	\kEq{\kappa}
+}{
+	\kEq{\klist{\kappa}}
+}
+\\
+\inferrule[u-eq]{ }{
+	\kEq{\kunit}
+}
+
+\inferrule[p-eq]{
+	\kEq{\kappa_1}\\
+	\kEq{\kappa_2}
+}{
+	\kEq{\kpair{\kappa_1}{\kappa_2}}
+}
+
+\inferrule[s-eq]{
+	\kEq{\kappa_1}\\
+	\kEq{\kappa_2}
+}{
+	\kEq{\ksum{\kappa_1}{\kappa_2}}
+}
+\end{mathpar}
+\caption{\small Kinding for type and operator constructors and external terms, and equality kinds (see text).
+Type-level terms stored in the constructor context are not needed during kinding, indicated using a dash.}
+\label{kindprog}
+\vspace{-10pt}
+\end{figure}
+\begin{figure}[t]
+\small
+%\vspace{-15pt}
+$\fbox{\inferrule{}{\tKindX{\tau}{\kappa}}}$
+%~~~$\chi ::= \cdot \pipe \checkmark$
+\begin{mathpar}
+\small\inferrule[k-var]{
+}{
+  \tKind{\tvarCtxX{t}{\kappa}}{\fvalCtx}{\tvar{t}}{\kappa}
+}
+~~~~
+\inferrule[k-arrow-i]{
+  \tKind{\tvarCtxX{t}{\kappa_1}}{\fvalCtx}{\tau}{\kappa_2}
+}{
+  \tKindX{\tlam{t}{\kappa_1}{\tau}}{\karrow{\kappa_1}{\kappa_2}}
+}
+~~~~
+\inferrule[k-arrow-e]{
+  \tKindX{\tau_1}{\karrow{\kappa_1}{\kappa_2}}\\\\
+  \tKindX{\tau_2}{\kappa_1}
+}{
+  \tKindX{\tapp{\tau_1}{\tau_2}}{\kappa_2}
+}
+
+\text{\color{gray} (kinding for integers, labels, lists, products and sums also standard)}
+
+%%
+%%\inferrule{ }{
+%%	\tKindX{\tstr{str}}{\kstr}
+%%}(\text{str}^I_\tau)
+%%
+%%\inferrule{ }{
+%%	\tKindX{\tunit}{\kunit}
+%%}(\text{1}^I_\tau)
+%%
+%%\inferrule{
+%%	\tKindX{\tau_1}{\kappa_1}\\
+%%	\tKindX{\tau_2}{\kappa_2}
+%%}{
+%%	\tKindX{\tpair{\tau_1}{\tau_2}}{\kpair{\kappa_1}{\kappa_2}}
+%%}({\times}^I_\tau)
+%%
+%%\inferrule{
+%%	\tKindX{\tau}{\kpair}{\kappa_1}{\kappa_2}
+%%}{
+%%	\tKindX{\tfst{\tau}}{\kappa_1}
+%%}({\times}^{E1}_\tau)
+%%
+%%\inferrule{
+%%	\tKindX{\tau}{\kpair}{\kappa_1}{\kappa_2}
+%%}{
+%%	\tKindX{\tsnd{\tau}}{\kappa_2}
+%%}({\times}^{E2}_\tau)
+%%\inferrule{ }{
+%%\inferrule{
+%%	\tKindX{\tau_{1}}{\kappa_{1}}\\
+%%	\tKindX{\tau_{2}}{\kappa_{2}}
+%%}{
+%%	\tKindX{\tpair{\tau_{1}}{\tau_{2}}}{\kpair{\kappa_{1}}{\kappa_{2}}}
+%%}~(\times_\tau)
+%%
+%%\inferrule{
+%%	\tKindX{\tau}{\kpair{\kappa_{1}}{\kappa_{2}}}
+%%}{
+%%	\tKindX{\tfst{\tau}}{\kappa_{1}}
+%%}~(\text{fst}_\tau)
+%%
+%%\inferrule{
+%%	\tKindX{\tau}{\kpair{\kappa_{1}}{\kappa_{2}}}
+%%}{
+%%	\tKindX{\tsnd{\tau}}{\kappa_{2}}
+%%}~(\text{snd}_\tau)
+%%
+%% TODO: list
+\inferrule[k-eq]{
+	\kEq{\kappa}\\
+	\tKindX{\tau_1}{\kappa}\\
+	\tKindX{\tau_2}{\kappa}\\\\
+	\tKindX{\tau_3}{\kappa'}\\
+	\tKindX{\tau_4}{\kappa'}
+}{
+	\tKindX{\tifeq{\tau_1}{\tau_2}{\kappa}{\tau_3}{\tau_4}}{\kappa'}
+}
+
+\inferrule[k-ty-i]{
+	\fval{Tycon}{\kappaidx}{-}{-} \in \fvalCtx\\\\
+	\tKind{\tvarCtx}{\fvalCtx}{\tauidx}{\kappaidx}
+}{
+	\tKindX{\ttype{Tycon}{\tauidx}}{\kTypeBlur}
+}
+
+\inferrule[k-ty-e]{
+	\tKindX{\tau}{\kTypeBlur}\\
+	\fval{Tycon}{\kappaidx}{-}{-} \in \fvalCtx\\\\
+	\tKind{\tvarCtxX{x}{\kappaidx}}{\fvalCtx}{\tau_1}{\kappa}\\
+	\tKindX{\tau_2}{\kappa}
+}{
+	\tKind{\tvarCtx}{\fvalCtx}{\tfamcase{\tau}{Tycon}{x}{\tau_1}{\tau_2}}{\kappa}
+}
+~~~~~~
+\inferrule[k-d-i]{
+	\tKindX{\tau}{\kTypeBlur}\\\\
+	\isItermX{\ibar}
+}{
+	\tKindX{\tden{\ibar}{\tau}}{\kDen}
+}
+
+\inferrule[k-d-i-checked]{
+	\tKind{\Delta}{\fvalCtx}{\tau}{\kTypeBlur}\\\\
+	\isItermX{\ibar}
+}{
+	\tKind{\Delta}{\fvalCtx}{\tden{\ibar}{\tau}^\checkmark}{\kDen}
+}
+~~~~~~~
+\inferrule[k-d-e]{
+	\tKindX{\tau}{\kDen}
+}{
+	\tKindX{\ttypeof{\tau}}{\kTypeBlur}
+}
+%
+%\inferrule[iterm-intro]{
+%	\isIterm{\tvarCtx}{\emptyctx}{\fvalCtx}{\iota}
+%}{
+%	\tKindX{\titerm{\iota}}{\kITerm}
+%}
+
+\inferrule[k-reptype-i]{
+	\isItype{\tvarCtx}{\fvalCtx}{\sbar}
+}{
+	\tKindX{\titype{\sbar}}{\kIType}
+}
+\end{mathpar}
+$\fbox{$\isItermX{\ibar}$}$
+\begin{mathpar}
+\inferrule[k-i-var]{ }{
+	\isItermX{\evar{x}}
+}
+~~~~~~
+\inferrule[k-i-fix]{
+	\isItypeX{\sbar}\\
+	\isItermX{\ibar}
+}{
+	\isItermX{\ifix{\evar{x}}{\sbar}{\ibar}}
+}
+~~~~~~
+\inferrule[k-i-transof]{
+	\tKindX{\tau}{\kDen}
+}{
+	\isItermX{\itransof{\tau}}
+}
+%\inferrule[i-lam-kinding]{
+%	\isItypeX{\sbar}\\\\
+%	\isItermX{\ibar}
+%}{
+%	\isItermX{\ilam{\evar{x}}{\sbar}{\ibar}}
+%}
+
+\text{\color{gray} (omitted rules are trivially recursive, like \textsc{k-i-fix})}
+%\inferrule[iterm-dereify-kinding]{
+%	\tKindX{\tau}{\kITerm}
+%}{
+%	\isItermX{\iup{\tau}}
+%}
+%
+\end{mathpar}
+$\fbox{$\isItypeX{\sbar}$}$
+\begin{mathpar}
+%\inferrule[i-int-kinding]{ }{
+%	\isItypeX{\dint}
+%}
+%
+%\inferrule[i-prod-kinding]{
+%	\isItypeX{\sigma_1}\\\\
+%	\isItypeX{\sigma_2}
+%}{
+%	\isItypeX{\dpair{\sigma_1}{\sigma_2}}
+%}
+%
+\inferrule[k-s-unquote]{
+	\tKindX{\tau}{\kIType}
+}{
+	\isItypeX{\dup{\tau}}
+}
+
+\inferrule[k-s-repof]{
+	\tKindX{\tau}{\kTypeBlur}
+}{
+	\isItypeX{\trepof{\tau}}
+}
+
+\text{\color{gray} (omitted forms are trivially recursive)}
+\end{mathpar}
+\caption{\small Kinding for type-level terms.
+The kinding context $\tvarCtx$ maps type variables to kinds.
+Checked derivates (introduced by rule \textsc{k-d-i-checked}, discussed in Sec.
+\ref{repcon}) should only be constructed by the semantics (this can be enforced by a parser, so we will assume it in the metatheory).}
+\label{tlkind}
+\vspace{-10px}
+\end{figure} 
+\begin{figure}[t]
+\small
+$\fbox{\inferrule{}{\ecompilesAX{e}{\tau}{\ibar}}}$
+\begin{mathpar}
+\inferrule[att-var]{ }{
+	\ecompilesA{\etCtxX{x}{\tau}}{\fvalCtx}{\evar{x}}{\tau}{\evar{x}}
+}
+
+\inferrule[att-lam]{
+	\tEvalX{\tau_1}{\tau_1'}\\
+%	\ddbar{\fvar{Arrow}}{\fvalCtx}{\trepof{\tau_1'}}{\sbar_1}\\
+	%\delfromtau{$\Xi_0$}{\fvalCtx}{\tau_1'}{\sabs}\\\\
+	\ecompilesA{\etCtxX{x}{\tau_1'}}{\fvalCtx}{e}{\tau_2}{\ibar}
+}{
+	\ecompilesA{\etCtx}{\fvalCtx}{\elam{\evar{x}}{\tau_1}{e}}{\ttype{Arrow}{\tpair{\tau_1'}{\tau_2}}}{\ilam{\evar{x}}{\trepof{\tau_1'}}{\ibar}}
+}
+
+\inferrule[att-op]{
+%	\tEvalX{\tauidx}{\tau_{\text{idx}}'}\\\\
+	\ecompilesA{\etCtx}{\fvalCtx}{e_1}{\tau_1}{\ibar_1}~~~~
+	\cdots~~~~
+	\ecompilesA{\etCtx}{\fvalCtx}{e_n}{\tau_n}{\ibar_n}\\
+	\fval{Tycon}{-}{-}{\theta} \in \fvalCtx\\
+	\tops{op}{\kappaidx}{i}{a}{\taudef} \in \theta\\
+	\tEvalX{\taudef~\tauidx~(\tden{\ibar_1}{\tau_1}^\checkmark :: \cdots :: \tden{\ibar_n}{\tau_n}^\checkmark :: \tnil{\kDen})}{\tinl{\kunit}{\tauD}}\\
+	\dcheck{\Gamma}{\fvalCtx}{\fvar{Arrow}, \fvar{Tycon}}{\tauD}{\tau}{\ibar}
+}{
+	\ecompilesA{\etCtx}{\fvalCtx}{\eop{Tycon}{op}{\tauidx}{\splat{e}{1}{n}}}{\tau}{\ibar}
+}
+\end{mathpar}
+\caption{\small The checked active typing judgement.
+The normalization judgement for type-level terms ($\Downarrow$) is specified in Fig.
+\ref{tleval}.
+Derivate checking is specified in Fig.
+\ref{ait}.
+Note that checked derivates can \emph{only} be constructed by the \textsc{att-op} rule (we enforce this judgementally in Sec.
+\ref{theory}).}
+\label{absatt}
+\end{figure}
+
 \section{Generic Functions, Targets and Interactive Execution}\label{targets}
@@
 
-\subsection{Metaprogramming in Ace}
-Generic functions can be generated directly from ASTs as well, providing Ace with support for  straightforward metaprogramming.
+%\subsection{Metaprogramming in Ace}
+%Generic functions can be generated directly from ASTs as well, providing Ace with support for  straightforward metaprogramming.
 Listing \ref{metaprogramming} shows how to generate two more generic functions, \verb|scale| and \verb|negate|.
@@
 \end{codelisting}
-Lines \ref{compscript}.3-\ref{compscript}.5 construct the types assigned to the arguments of \verb|map| on lines \ref{compscript}.7-\ref{compscript}.10.
+%Lines \ref{compscript}.3-\ref{compscript}.5 construct the types assigned to the arguments of \verb|map| on lines \ref{compscript}.7-\ref{compscript}.10.
 In Ace, types are themselves values that can be manipulated at compile-time.
@@
 \ref{atypes}).
-As Python values, types can be assigned to variables when convenient (removing the need for  facilities like \verb|typedef| in C or \verb|type| in Haskell).
+%As Python values, types can be assigned to variables when convenient (removing the need for  facilities like \verb|typedef| in C or \verb|type| in Haskell).
 Types, like all compile-time objects derived from Ace base classes, do not have visible state and operate in a referentially transparent manner (by constructor memoization, which we do not detail here).% These types are all implemented in the \verb|clx| library imported on line 1, none are built into Ace itself.

### 2014-03-26 ace-oopsla14/string-appendix.tex
Content finished; starting cleanup.

--- 2014-03-25 ace-oopsla14/string-appendix.tex
@@
 \begin{document}
-
-TODO-nrf spell check.
-
-TODO-nrf Include functions (app, abs)?
-
-TODO-nrf Change icheck in eval rule to Python.
-
-TODO-nrf Is it ok to make the proofs single-column? If not clean up overflows.
-
-TODO-nrf Make sure failure of unicity isn't causing any problems and write down lemmas about
-forms (e.g. $e \in \lang{r} \implies e = \strf{s}$ etc).
 
@@
 
+\subsection{Typing}
+
+The $\strin[r]$ type is parameterized by regular expressions; if $e:\strin[r]$,
+then $e \in r$. Mapping from an arbitrary $\str$ to a $\strin[r]$ requires
+defining a algorithm -- in terms of filter -- for converting a $\strin[.*]$
+into a $\strin[r]$. The static semantics of the language defines the types of
+operations on regular expressions in terms of well-understood properties about
+regular lanugages; we recall these properties in section 3.
+
+\subsection{Dynamics}
+
+There are two evaluation judgements: $e:T \reduces e'$ and $e:T \ireduces i$.
+The $\reduces$ relation is between $\lcs$ expressions, while the $\ireduces$
+relation is a mapping from $\lcs$ expressions into internal language expressions
+$i$ such that $i \ival$.
+
+Safety of the evaluation relation depends upon an injective mapping from $\lcs$ types info
+internal language types. This relation, $h$, is defined below.
+
 \onecolumn
@@
 
-\inferline{E-concat}
+\inferline{E-concatval}
 {e_1 + e_2 : \strin[r_1+r_2] \ireduces \istrf{e_1} + \istrf{e_2}}
-{ }
+{e_1 \val \\ e_2 \val }
+
+\inferline{E-concatL}
+{e_1 + e_2 : \strin[r_1+r_2] \reduces {e_1}' + {e_2}}
+{ e_1 : \strin[r_1] \reduces {e_1}'}
+
+\inferline{E-concatR}
+{e_1 + e_2 : \strin[r_1+r_2] \reduces {e_1} + {e_2}'}
+{ e_2 : \strin[r_2] \reduces {e_2}'}
 
@@
 \inferline{E-convertval}
+{ [\strin[r']](e) : \strin[r'] \ireduces \istrf{e} }
+{ e \val }
+
+\inferline{E-convert}
 { [\strin[r']](e) : \strin[r'] \reduces [\strin[r']](e') }
@@
 
-\inferline{E-convert}
-{ [\strin[r']](e) : \strin[r'] \ireduces \istrf{e} }
-{ e \val }
-
-\inferline{E-dconvertval (optional) \footnote{See definition ???} }
+\inferline{E-dconvertval (optional) }
 {\dconvert{\strin[r']}{e} : \strin[r'] \reduces {\dconvert{\strin[r']}{e'}}}
@@
 
-\inferline{E-dconvert (optional) \footnote{See definition ???} }
+\inferline{E-dconvert (optional) }
 {\dconvert{\strin[r']}{e} : \strin[r'] \ireduces \ifilter{r'}{\istrf{e}}}
@@
 \end{tabular}
-\caption{Operational semantics for $\lcs$}
+\caption{SOS rules for $\lcs$}
 \label{fig:lcsEval}
@@
 \end{figure}
+
 \twocolumn
 
-\subsection{Typing}
+\section{Type Safety}
 
-The $\strin[r]$ type is parameterized by regular expressions; if $e:\strin[r]$,
-then $e \in r$. Mapping from an arbitrary $\str$ to a $\strin[r]$ requires
-defining a algorithm -- in terms of filter -- for converting a $\strin[.*]$
-into a $\strin[r]$. The static semantics of the language defines the types of
-operations on regular expressions in terms of well-understood properties about
-regular lanugages; we recall these properties in section 3.
+The type safety proof relies upon some assumptions about the type system and
+dynamics of the internal language, as well as some properties of regular
+languages.
 
-\subsection{Dynamics}
-
-There are two evaluation judgements: $e:T \reduces e'$ and $e:T \ireduces i$.
-The $\reduces$ relation is between $\lcs$ expressions, while the $\ireduces$
-relation is a mapping from $\lcs$ expressions into internal language expressions
-$i$ such that $i \ival$.
-
-Safety of the evaluation relation depends upon an injective mapping from $\lcs$ types info
-internal language types. This relation, $h$, is defined below.
+There must exist a translation from $\lcs$ types to the types of the internal
+language. For the remainder of this paper, we call the type translation function
+$h$.
 
@@
 
-\section{Type Safety}
-
-The type safety proof relies upon some established results about regular languages.
-The important definitions and theorems follow.
-
-\subsection{Properties of Regular Languages}
- % copy from thesis
-
-\subsection{Type Safety Proof}
-
-Before presenting progress and preservation theorems, we establish some definitions
-and lemmas necessary for the proofs.
+Additionally, we assume that the internal language contains an implementation of strings,
+together with operations for concatenation and filtering by regular expression.
 
 \begin{defn}[Types of internal values]
-The types of internal values are defined as follows:
+  Let $\istrf{s}$ range over string literals and $r$ over regular expressions.
+  Internal values are typed as follows:
 \begin{itemize}
   \item If $e=\istrf{s}$ then $e:\istr$.
-  \item If $e=\ifilter(r, \istrf{s})$ then $e:\istr$.
+  \item If $e=\ifilter{r}{\istrf{s}}$ then $e:\istr$.
   \item If $e=\istrf{s_1}+\istrf{s_2}$ then $e:\istr$.
@@
 
-Definition 2 is an additional assumption about
-the internal language. To summarize, we assume an internal language containing 
-strings together with operations for string concatentation and filtering. We
+For simplicity, we assume a fixed translation from
+$\lcs$ regular expressions to regular expressions recognizable by the internal
+language's regex library (in practice, a fixed translation is acceptable.)
+To summarize, we assume an internal language containing 
+a string typetogether with operations for string concatentation and filtering. We
 expect closure over strings for both operations.
-Recall that $\ifilter$ is only needed for dynamic casts, which may be removed without
+Finally, recall that $\ifilter$ is only needed for dynamic casts, which may be removed without
 descreasing the expressivity or even usability of the language.
+Finally, the semantics of the filter function are defined in terms of ${\tt rl\_filter}$,
+which is a static version of ${\tt ifilter}$.
+
+\subsection{Properties of Regular Languages}
+
+The regular languages are the smallest set generated by regular expressions
+defined in Figure 1.
+
+\begin{thm}{Closure Properties.} \label{thm:closure}
+The regular expressions are closed under complements and concatenation.
+\end{thm}
+\begin{proof}
+See \cite{cinderella}.
+\end{proof}
+
+\begin{thm}{Coercion Theorem.} \label{thm:coerce}
+Suppose that $R$ and $L$ are regular expressions, and that $s \in R$ is a finite string.
+Let $s' := coerce(R,L,s)$ with all maximal substrings recognized by $L$ replaced with $\epsilon$.
+Then $s'$ is recognized by $(R \backslash L) + \emptyset$ and the construction of $R \backslash L$ is decidable.
+\end{thm}
+\begin{proof}
+Let $F,G$ be FAs corresponding to $R$ and $L$, and let $G'$ be $G$ with its final states inverted (so that $G'$ is the complement of $L$).
+Define an FA $H$ as a DFA corresponding to the NFA found by combining $F$ and $G'$ such that $H$ accepts only if $R$ and $L'$ accept or if $s$ is empty (this construction may result in an exponential blowup in state size.)  Clearly, $H$ corresponds to $R \backslash L + \emptyset$.
+Thus, the construction of $R \backslash L + \emptyset$ is decidable.
+
+If $R \subset L$, $s' = \emptyset$.
+If $L \subset R$, either $s' = \emptyset$, or $s' \in R$ and $s' \not \in L$.
+If $R$ and $L$ are not subsets of one another, then it may be the case that $L$ recognizes part of $R$.
+Consider $L$ as the union of two languages, one which is a subset of $R$ and one which is disjoint.
+The subset language is considered above and the disjoint language is inconsequential.
+\end{proof}
+
+\subsection{Type Safety Proof}
 
@@
   \begin{itemize}
-    \item If $e:T$ and $e:T \ireduces i$ then $i : \sigma$ for some $\sigma$.
+    \item If $e:T$ and $e:T \ireduces i$ then $i : \sigma$ such that $h(T) = \sigma$.
     \item If $e:T$ and $e:T \reduces e'$ then $e':T$.
@@
 It suffices to show that $h(\strin[r]) = \istr$, which follows from the definition of $h$ for arbitrary $r$.
-  \item \textbf{E-concat}.
+  \item \textbf{E-concatval}.
 By the definition of typing for internal terms, $\istrf{e_1} + \istrf{e_2} : \istr$.
 So it suffices to show that $h(\strin[r_1+r_2]) = \istr$, which follows from the definition of $h$ for arbitrary $r_1$, $r_2$.
+  \item \textbf{E-concatR, E-concatL}.
+WLOG consider E-concatR.
+By induction, $e_1' : \strin{l_1}$.
+By inversion of T-Concat at the premise, $e_2 : \strin{r_2}$.
+Therefore, $e_1 + e_2 : \strin[r_1 + r_2]$.
   \item \textbf{E-Filterval}. We have that $\filter{\strin[r']}{e}:\strin[r\backslash r' + \emptyset]$. By inversion of T-Filter, $e:\strin[r]$.
     By T-Equiv-string\_in (which is bidirectional), $e \in \lang{r}$.
-    By theorem 2, ${\tt rl\_filter}(r,e) \in \lang{r\backslash r' + \emptyset}$.
+    By the Coercion Theorem, ${\tt rl\_filter}(r,e) \in \lang{r\backslash r' + \emptyset}$.
     By T-Equiv-string\_in,  $e \in \lang{r}$ and ${\tt rl\_filter}(r,e) \in \lang{r\backslash r' + \emptyset}$ implies ${\tt rl\_filter}(r,e):\strin[r\backslash r' + \emptyset]$.
-  \item \textbf{E-Filter}. By inversion, $e : \strin[r'] \reduces e'$ so by the induction $e':\strin[r']$.
-    Therefore, ${\tt filter}(\strin[r], e') : \strin[r \backslash r' + \emptyset]$ by T-Filter.
+  \item \textbf{E-Filter}. By inversion, $e : \strin[r']$ $\reduces e'$ so by the induction $e':\strin[r']$.
+    Therefore, \\${\tt filter}(\strin[r], e'):$ $\strin[r \backslash r' + \emptyset]$ by T-Filter.
   \item \textbf{E-Convertval}. It suffices to show that $h(\strin[r]) = \istr$, which is true by definition.
@@
 By induction on the derivation of $e:T$.
-\begin{itemize}[label=$ $,itemsep=1ex]
-  \item \textbf{T-Equiv-Include}. Since $e \in \lang{r}$, $e = \strf{s}$ for some $s$; therefore, $e \val$ by E-Strinval.
-  \item \textbf{T-Minimum}. 
-\end{itemize}
+For \textbf{T-Equiv-Include}, note that $e \in \lang{r}$, $e = \strf{s}$ for some $s$; therefore, $e \val$ by E-Strinval. 
+We have that $e \val$ and $e : \strin[r]$, so $e \ireduces \istrf{e}$ by E-string\_in.
+The remaining cases follow by induction on the hypotheses and application of corresponding evaluation rules.
 \end{proof}
@@
 
+The $\lcs$ language is implemented as a type system extension in Ace; this extension
+is illustrated in the examples at the beginning of this paper.
+
+Computing a regular expression representating the language
+$R \backslash S$ is necessary in order to type-check expressions in which the filter function
+occurs. This language is computed by translating $R$ and $S$ into finite automata,
+complementing the final states of $S$, then constructing the cross-product of $R$
+and $S'$. Type checking terminates only because this construction is decidable.
+
+In addition to this construction, other more typical operations -- such
+as equality checks for regular expressions and regular expression matching -- are
+also necessary. For these reasons, the Ace extension implementing $\lcs$ includes a library
+implementing each of these constructions.
+
 \end{document}

### 2014-03-26 ace-oopsla14/string-appendix.tex
String proof reading for grammar etc.

@@
 SQL injection attacks. To prevent such attacks, we might ensure that any string
-used as input to a query does not contain unescaped SQL common sequences.
+used as input to a query does not contain unescaped SQL command sequences.
 
-This appendix presents a type system called $\lcs$ for ensuring that input 
+This appendix presents a type system, $\lcs$, for ensuring that input 
 sanitation algorithms are implemented correctly with respect to use site 
@@
 Ruby, our type system provides a \emph{static} guarantee that input is always 
-properly sanitized before use. We achieve this by moving portions of the regular
-expression library into the type system. Therefore, type safety relies upon 
+properly sanitized before use. We achieve this by defining a typing relation
+which captures idiomatic sanitation algorithms. Type safety relies upon 
 several closure and decidability results about regular languages.
 
-Although Ur/Web achieves a similar safety guarantee, its type system is significantly
-more complicated than the system defined below. We believe this simplicity 
-demonstrates the power of extensibility. Instead of introducing new abstractions
-based upon subtle analyses and requiring the adoption of new programming languages, 
-extension designers may capture natural programming idioms directly. TODO-nrf this seems unsubstantiated or dangerously vague.
-
-Finally, XDuce typechecks XML. Like our work, XDuce relies upon some properties of
+XDuce typechecks XML. Like our work, XDuce relies upon some properties of
 regular expression to establish soundness and copmleteness results. Unlike
@@
 strings (as opposed to tree-structured XML documents). Furthermore, our static
-treatment of the ubiquitous ``filter" function for regular expression is novel.
+treatment of the ubiquitous regex filter (replace) function for regular expression is novel.
+
+Finally, some research languages achieve similar safety guarantees via a specialized type
+systems. The existence of such domain-specific type systems is suggestive, and 
+we believe the simplicity of $\lcs$ demonstrates the effectiveness of
+extensibility. Instead of introducing entirely new languages for each domain,
+language extension developers may invest incrementally in obtaining static
+guarnatees which address common mistakes.
 

### 2014-03-26 ace-oopsla14/ace-oopsla14.tex
OOPSLA14 final

@@
 \usepackage{listings}
-
+\usepackage{wasysym}
     \makeatletter
@@
   basicstyle=\ttfamily\scriptsize,
-  morekeywords={lambda, self, assert, as,type},
+  morekeywords={lambda, self, assert, as, cls},
   numbers=left,
@@
 
-\title{\Ace: An Extensible Static Type System Inside a Python}
+\title{\Ace: Active Typechecking and Translation Inside a Python}
 
@@
 In addition to describing a full-scale language design, we give a simplified calculus that describes the foundations of this mechanism and show that, as implemented in Ace, it is flexible enough to admit practical library-based implementations of types drawn from functional, object-oriented, parallel and domain-specific languages.
-We also show how to construct safe, natural and extensible foreign function interfaces using a novel form of ``staged'' type propagation from Python into Ace and then through to OpenCL.
+%We also show how to construct safe, natural and extensible foreign function interfaces using a novel form of ``staged'' type propagation from Python into Ace and then through to OpenCL.
 % can be used to enable type system extensions and demonstrates how type safety can be maintained by lifting a technique from the typed compilation literature into the language.%To show that the method scales, we give a full implementation of the OpenCL programming language as a library.
@@
 Python itself, as well as the others just described) can also be compiled and invoked interactively from Python.
-Ace supports a form of staged \emph{ad hoc} polymorphism (based on singleton types) wherein the ``dynamic type'' of a Python value can determine a static type and trigger active typechecking and translation ``just-in-time'' when Python code calls a generic Ace function.
+%Ace supports a form of staged \emph{ad hoc} polymorphism (based on singleton types) wherein the ``dynamic type'' of a Python value can determine a static type and trigger active typechecking and translation ``just-in-time'' when Python code calls a generic Ace function.
 We show how this can significantly streamline interactive workflows (e.g.
@@
 \ref{discussion} by summarizing the key features needed by a host language to support active typechecking and translation, and discussing present limitations and potential future work.
-The appendix has details omitted for concision as well as several more examples, including functional datatypes, flexible complex numbers and a previously little-explored type system for tracking string invariants using regular expressions statically (which we show being used in Listing \ref{example}, amongst  other examples).
+%The appendix has details omitted for concision as well as several more examples, including functional datatypes, flexible complex numbers and a previously little-explored type system for tracking string invariants using regular expressions statically (which we show being used in Listing \ref{example}, amongst  other examples).
 
@@
 The field \verb|name| has type \verb|string|, defined in \verb|examples.py|, while the fields \verb|account_num| and \verb|routing_num| have more interesting types, classifying strings guaranteed statically to be in a regular language specified statically by a  regular expression pattern (written, to avoid needing to escape backslashes, using Python's \emph{raw string literals}).
-The appendix specifies the typechecking rules, based on \cite{fulton-thesis}.
+The appendix will give the typechecking rules, based on \cite{fulton-thesis}.
 To our knowledge, this type system has not previously been implemented.
-We include it to emphasize that we aim to support specialized type systems,  not just general purpose constructs like records, numbers and strings.
+We include it to emphasize that we aim to support specialized type systems,  not just general purpose constructs like records and strings.
 \item 
@@
 On line 14, we construct a simple \emph{prototypic object type} \cite{Lie86}.
-The type \verb|Transfer| classifies terms consisting of a \todo{what is the right term for this?}\emph{fore} of type \verb|A| and a \emph{prototype} of type \verb|C|.
+The type \verb|Transfer| classifies terms consisting of a \emph{fore} of type \verb|A| and a \emph{prototype} of type \verb|C|.
 If a field cannot be found in the fore, the type system will delegate (here, statically) to the prototype.
@@
 As a result, we can assume on line 19 that \verb|t.amount| and \verb|t.name| have types \verb|decimal[2]| and \verb|string|, respectively.
-We will return to the details of \verb|print| and the form \verb|[string](t.amount)|, which performs an explicit conversion, shortly.\todo{when?}
+We will return to the details of \verb|print| and the form \verb|[string](t.amount)|, which performs an explicit conversion, in Sec.
+\ref{att}.
 
@@
 def print_transfer(t):
-  print "Transferring %s to %s." % (
-    __ace_0.decimal_to_str(t[0],2), t[1][0])
+  print ("Transferring %s to %s." % (
+    __ace_0.decimal_to_str(t[0],2), t[1][0]))
 
@@
 To typecheck, translate and execute the code in Listing \ref{example}, we have two choices: do so externally at the shell, or perform compilation interactively (or implicitly) from within Python.
-We will begin with the former, then return to interactive and implicit compilation in Sec.
-\ref{targets}.% with our OpenCL examples.
+We will begin with the former.
+Interactive and implicit compilation are implemented but we do lack space to detail it in the present paper, choosing here to focus on the core mechanism.% with our OpenCL examples.
 
@@
 \ref{att}).
-If no type errors are discovered, the translations are collected (obeying order dependencies) and emitted, with file extension(s) determined by the target(s) in use, discussed in Sec.
-\ref{targets}.
+If no type errors are discovered, the translations are collected (obeying order dependencies) and emitted, with file extension(s) determined by the target(s) in use, discussed further in Sec.
+\ref{att}.
 Here our target is  the default target associated with the base \verb|py|, which emits Python 2.6 files.
@@
 \begin{lstlisting}
-from listing1 import py, A
+from listing1 import py, A, record
 @py
@@
 It can be used as a decorator because \verb|ace.Base| overloads the call operator, shown in Listing \ref{base}.
-The \verb|_process| function (not shown) \todo{show in appendix?}operates as follows:
+The \verb|_process| function (not shown) operates as follows:
 \begin{enumerate}
@@
 \subsection{Checking Typed Functions}
-When the Ace compiler begins typechecking a \verb|TypedFn| (when asked to by \verb|acec|, or when it is  invoked interactively as we will show in Sec. \ref{targets}), it proceeds as follows:
+When the Ace compiler begins typechecking a \verb|TypedFn| (when asked to by \verb|acec|, or when it is  invoked interactively, not discussed in this paper), it proceeds as follows:
 \begin{enumerate}
@@
 This can be seen on lines 2-4 of Listing \ref{pybase}, where the \verb|PyBase| base adds an attribute tracking local bindings (initially only the arguments) and the return type, which will be synthesized but initially is not known.
-\todo{briefly mention return type annotations?}
-\item For each statement in the body (after the documentation string and type annotation, if present), the compiler calls the context's \verb|check| method, which in turn defers to a method of the base, \verb|check_|$F$, where $F$ is the form of the statement (taken directly from Python's \verb|ast| package \cite{python}).
-These methods are responsible for checking that the statement is well-typed (raising an \verb|ace.TypeError| if not) and having any necessary effect on the context.
+\item For each statement in the body (after the documentation string and type annotation), the compiler delegates to either the base, or of a type synthesized from a subterm, by calling a method named \verb|check_|$F$, where $F$ is the form of the statement (derived directly from Python's \verb|ast| package \cite{python}).
+These methods are responsible for checking that the statement is well-typed (raising an \verb|ace.TypeError| if not) and having any needed effect on the context.
 \item The base synthesizes a type for the function as a whole via the \verb|syn_FunctionDef_outer| method.
@@
 \end{enumerate}
-
-Because bases are user-defined objects, this represents the first instance of the compiler delegating control to a library.
+The assignment statements on lines \ref{att-example}.5\footnote{In this section, we will need to refer to code in Listings \ref{att-example}, \ref{pybase}, \ref{context} and \ref{record}, so we adopt this syntactic convention to refer to line numbers for concision.} and \ref{att-example}.6 are checked by the base method \verb|check_Assign_Name| shown on lines \ref{pybase}.6-\ref{pybase}.14 and the return statement on line \ref{att-example}.7 is checked by the base's \verb|check_Return| on lines \ref{pybase}.19-\ref{pybase}.23.
+Both work similarly: if this is the first assignment to a particular name, or the first return statement seen, then the value must be able to synthesize a type in the current context.
+Otherwise, it is analyzed against the previously synthesized type.
+
 \subsection{Bidirectional Typechecking}
@@
 
-  def check_Assign(self, ctx, s):
-    ace.util.unsupported_if(len(s.targets) != 1
-      or not isinstance(s.targets[0], ast.Name))
-    x, e = s.targets[0].id, s.value # see ast module
+  def check_Assign_Name(self, ctx, s):
+    x, e = s.target.id, s.value
     if x in ctx.locals:
@@
 
-  def trans_Assign(self, ctx, target, s):
+  def trans_Assign_Name(self, ctx, target, s):
     return target.direct_translation(ctx, s)
@@
       return self.trans_lifted(ctx, target, e)
+
+  default_Dict_asc = default_Str_asc = dyn
   # ...
-  default_target = PyTarget()   
-  default_Dict_asc = dyn
-  # ...
+  def init_target(ctx): return PyTarget()   
 py = PyBase()
@@
     self.fn = fn
-
-  def check(self, s):
-    if isinstance(s, ast.Assign):
-      self.fn.base.check_Assign(self, s)
-    # ... other statement forms similar
 
@@
     # ... other compound forms similar (cf appendix)
-    elif isinstance(e, ast.Dict):
+    elif isinstance(e, ast.Str):
       return self.ana(self, e, 
-        self.fn.base.default_Dict_asc)
+        self.fn.base.default_Str_asc)
     # ... other unascribed literal forms similar
@@
 
-  def __eq__(self, other): 
-    if self.anon or other.anon: 
-      return self is other
-    else: 
-      return (isinstance(other, record) 
-        and self.sig.equiv(other.sig))
+  @classmethod 
+  def syn_Dict(cls, ctx, e):
+    sig = Sig((f, ctx.syn_ty(v)) 
+      for f, v in zip(e.keys, e.values))
+    return cls(sig, anon=True)
 
@@
       else:
-        raise ace.TypeError("...bad field...", f)
+        raise ace.TypeError("...extra field...", f)
     if len(self.sig) != len(e.keys):
       raise ace.TypeError("...missing field...", e)
-
-  @classmethod # (when ascription is a type constructor)
-  def syn_Dict(cls, ctx, e):
-    sig = Sig()
-    for f, v in zip(e.keys, e.values):
-      sig[f] = ctx.syn_ty(v)
-    return cls(sig, anon=True)
 
@@
         ctx.trans(target, e.value), target.Num(idx))
+
+  def __eq__(self, other):
+    if isinstance(other, record): 
+      if self.anon or other.anon: return self is other
+      else: return self.sig == other.sig
+
+  def trans_type(self, target):
+    return target.dyn
 \end{lstlisting}
@@
 \end{codelisting}
-The assignment statements on lines \ref{att-example}.5\footnote{In this section, we will need to refer to code in Listings \ref{att-example}, \ref{pybase}, \ref{context} and \ref{record}, so we adopt this syntactic convention to refer to line numbers for concision.} and \ref{att-example}.6 trigger the method \verb|check_Assign| \todo{if time, add destructuring assignment to appendix}shown on lines 6-14 of Listing \ref{pybase} and the return statement on line 7 triggers the method \verb|check_Return| on lines 18-22.
-Both work similarly: if this is the first assignment to a particular identifier, or the first return statement seen, then the value must synthesize a type in the current context.
-Otherwise, it is analyzed against the previously synthesized type in the current context.
+ Synthesis and analysis are mediated by the context via its \verb|syn| and \verb|ana| methods.
+To see how it works, let us begin at the first statement in our example, the assignment statement on line \ref{att-example}.5.
+As just stated, the method \verb|check_Assign_Name| on lines \ref{pybase}.6-\ref{pybase}.14 is called.
+Because the locals dictionary added to the context by the base does not yet have a binding for \verb|x|, the base asks to synthesize a type for the value being assigned, \verb|a.amount|, by calling \verb|ctx.syn|.
 
@@
 
-The method \verb|check_Assign| on lines \ref{pybase}.6-\ref{pybase}.14 is called.
-Because the locals dictionary in \verb|ctx| does not yet know what type \verb|x| should be, the base asks to synthesize a type for the expression being bound, \verb|a.amount|, by calling \verb|ctx.syn|.
-
 %\paragraph{Active Type Synthesis} 
 This method is defined on lines \ref{context}.5-\ref{context}.26.
-The relevant case of the conditional is the one on line \ref{context}.9, because \verb|a.amount| is of the form \verb|Attribute| according to Python's grammar.
-Control over this term is given to the type recursively synthesized for its value, \verb|a|.
-We recurse back into \verb|syn|, now taking the first branch, corresponding to terms of the form \verb|ast.Name|.
-Control over names is given to the base, as seen on line \ref{context}.7, by calling the \verb|syn_Name| method.
-We can see its implementation for the base we are using on lines \ref{pybase}.39-\ref{pybase}.46.
+The relevant case in this method is the one on line \ref{context}.9, because \verb|a.amount| is of the form \verb|Attribute| according to Python's grammar.
+The context delegates to the type recursively synthesized for its value, \verb|a|.
+We recurse back into \verb|syn|, now taking the first branch for terms of the form \verb|ast.Name|.
+Synthesizing a type for a name is delegated to the base by calling its \verb|syn_Name| method.
+We can see its implementation for the base we are using on lines \ref{pybase}.37-\ref{pybase}.44.
 The identifier \verb|a| is an argument to the function, which the base included in the initial locals dictionary, so we hit a base case and the type \verb|A| is synthesized.
 
-We can now pop back up to line \ref{context}.11, which can now delegates control over \verb|a.amount| to the type \verb|A|.
+We can now pop back up to line \ref{context}.11, which can now delegates synthesis of a type for \verb|a.amount| to the type \verb|A| via the \verb|syn_Attribute| method.
 Recall that \verb|A| was constructed using the \verb|record| constructor in Listing \ref{example}.
@@
 The class decorator \verb|slices_to_sig| provides the convenient notation shown used in Listing \ref{example} (by adding a metaclass to override the class object's subscript operator, not shown).
-The \verb|syn_Attribute| method is shown on lines \ref{record}.38-\ref{record}.42.
-It simply looks in the record signature for the provided attribute, so \verb|decimal[2]| is synthesized.
-We can now go back up to the assignment statement that triggered this chain of calls and see on line \ref{pybase}.14 that a binding for \verb|x| is added to the locals dictionary and checking of this statement thus succeeds (no \verb|ace.TypeError| was raised).
-
-The next statement also assigns to \verb|x|, but this time, the analysis branch is taken using the \verb|ana| method of \verb|Context|, shown on lines \ref{context}.28-\ref{context}.37.
-In Ace, analysis is only used for literal forms.
+The \verb|syn_Attribute| method is shown on lines \ref{record}.30-\ref{record}.34.
+It simply looks in the signature for the provided field name, so \verb|decimal[2]| is synthesized.
+We can now go back up to the assignment statement that triggered this chain of calls and see on line \ref{pybase}.14 that a binding for \verb|x| is added to the locals dictionary and checking of this statement succeeds.
+
+The next statement also assigns to \verb|x|, but this time, the base asks to analyze the value against \verb|A| using the \verb|ana| method of \verb|Context|, shown on lines \ref{context}.28-\ref{context}.37.
+Analysis differs from synthesis only for unascribed literal forms.
 For any other form, the context simply asks for an equal type to be synthesized (we will discuss in Sec.
 \ref{discussion} future work on integrating subtyping, where this would be relaxed).
-For concision, we leave the details of synthesis for binary operators to the appendix, which also shows the definition of the \verb|decimal| type constructor.\todo{do this based on cplx; change intro}
+For concision, we leave the details of the delegation protocol for binary operators to the appendix, which also shows the definition of the \verb|decimal| type constructor.
+The addition of two terms of type \verb|decimal[2]| has type \verb|decimal[2]|, so checking the second assignment statement also succeeds.
 
@@
 Because it is the first return statement encountered, it too requires that the returned value synthesize a type.
-Here the returned value is an ascribed literal.
+Here it is an ascribed literal.
 We can see on lines \ref{context}.17-\ref{context}.22 how this is handled.
-Because the ascription is a type, \verb|A|, rather than a type constructor, the literal is analyzed against that type.
-Lines \ref{context}.29-\ref{context}.30 show that this involves simply calling the type's \verb|ana_Dict| method, shown on lines \ref{record}.7-\ref{record}.14.
-This simply ensures that all and only the fields in the signature were provided, and analyzes the values against the corresponding type from the signature.
-
+Because the ascription is a type constructor, \verb|record|, rather than a type, the literal synthesizes a type by calling a \emph{class method}, \verb|record.syn_Dict|, shown on lines \ref{record}.6-\ref{record}.10.
+This method constructs a record signature by synthesizing a type for each value (using comprehension syntax for concision) then returns a new instance of the class.
+It is marked as anonymous.
+Anonymous records differ only in how equality is decided, shown on lines \ref{record}.44-\ref{record}.47.
+We want anonymous records to be equal structurally, while records declared like \verb|A| are by default distinct even if they have the same signature (following the convention of most functional languages with record declarations).
+
+Synthesizing a type for the value of the field \verb|y|, also an ascribed literal, takes a different path because the ascription is a type, \verb|A|, not a type constructor.
+Type ascriptions, as we have seen in the previous examples, cause the literal to analyzed.
+This proceeds through the \verb|ana_Dict| method on lines \ref{record}.12-\ref{record}.19, which analyzes the value of each field against the corresponding type in the signature, raising a type error if there are missing or extra fields.
+%This succeeds because \verb|x| analyzes against \verb|decimal[2]| because the base synthesizes that type, as described above.
+
+Synthesizing a type for the value of the field \verb|remark|, an unascribed literal form, follows a third path, seen on lines \ref{context}.13-\ref{context}.15.
+Unascribed literals are treated as if they had been  given the default ascription specified by the base.
+Our base specifies the default ascription as \verb|dyn| on line \ref{pybase}.52, which accepts our  literal, so synthesis succeeds.
+The return type of \verb|example| is thus equal to:
+\begin{lstlisting}[numbers=none]
+record(Sig(('y', A), ('remark', dyn)), anon=True)
+\end{lstlisting}
+\vspace{-10px}
 \subsection{Active Translation}
-Once typechecking is complete, the compiler enters the translation phase, generating terms in the target language specified by the base\todo{should we separate them?}.
-This phase follows essentially an identical protocol, using \verb|trans_|$X$ methods paired with each \verb|check_|/\verb|syn_|/\verb|ana_|$X$  method.
-The typechecking phase saves the entity that was delegated control as an attribute, \verb|delegate|, as well as the type of each AST node, \verb|ty|, so that these need not be determined again.
-The protocol for expressions can be seen on lines \ref{context}.39-\ref{context}.47.
-
-Translation methods have access to the context and node just as during typechecking, as well as an object, \verb|target|, that mediates code generation (named \verb|target| here).
-Because we are targeting Python directly, most of the \verb|trans| methods implement direct translations (factored out into a helper function).
-The main methods of interest are \verb|trans_FunctionDef_outer| on lines \ref{pybase}.33-\ref{pybase}-37 and the two translation methods in Listing \ref{record}, which implement the logic described in Sec.
+Once typechecking is complete, the compiler enters the translation phase.
+The base creates an instance of a class inheriting from \verb|ace.Target| for use during this phase via the \verb|init_target| method (line \ref{pybase}.54).
+This object provides methods for code generation and supports features like fresh variable generation, adding imports and so on.
+It's interface is not constrained by Ace (we will see what Ace requires of a target below) and the mechanics of code generation are orthogonal to the  focus of this paper, so we will discuss it relatively abstractly.
+The simplest API would be string-based code generation.
+For the Python target we use here, we generate ASTs.
+The API is based directly on the \verb|ast| library, with a few additional conveniences just mentioned.
+
+This phase follows the same delegation protocol as the typechecking phase.
+Each \verb|check_|/\verb|syn_|/\verb|ana_|$X$  method has a corresponding \verb|trans_|$X$ method.
+The typechecking phase saved the entity that was delegated control, along with the type assignment, as attributes of each node, \verb|delegate| and \verb|ty| respectively, so that these need not be determined again.
+This protocol can be seen on lines \ref{context}.39-\ref{context}.47.
+Translation methods have access to the context and node, as during typechecking, as well as the target.
+
+Because we are targeting Python directly, most of our \verb|trans| methods are direct translations (factored out into a helper function).
+The main methods of interest that are not entirely trivial are \verb|trans_FunctionDef_outer| on lines \ref{pybase}.33-\ref{pybase}-37 and the two translation methods in Listing \ref{record}, which implement the logic described in Sec.
 \ref{compilation}.
 
+Each type constructor must also specify a \verb|trans_type| method.
+This is important when targeting a typed language (so that the translations of type annotations can be generated, for example).
+As we will see in the next two sections, this is also critical to ensuring type safety.
+The language \emph{checks} not only that translations having a given type are well-typed, but that they have the type specified by this method (requiring that the target provide a \verb|is_of_type| method).
+Here, because we are simply targeting Python directly, the \verb|trans_type| method on line \ref{record}.49-\ref{record}.50 simply generates \verb|target.dyn|.
+
+When this phase is complete, each node processed by the context will have a translation, available via the \verb|trans| attribute.
+In particular, each typed function has a translation.
+Note that some nodes are never processed by the context because they were reinterpreted by the delegate (e.g.
+the field names in a record literal), so they do not have translations (as expected, given our discussion above).
+
+To support external compilation, the target must have an \verb|emit| method that takes the compilation script's file name and a reference to a string generator (an instance of \verb|ace.util.CG|) and emits source code.
+The string generator we provide can track indentation levels (to support Python code generation and make generation for other languages more readable, for the purposes of debugging).
+It allows non-local string generation via the concept of user-defined \emph{locations}.
+Each file that needs to be generated is a location and there can also be locations within a file (e.g.
+the imports vs.
+the top-level code), specified by a target the first time it finds that a necessary location is not defined.
+A generated entity (e.g.
+an import, class definition or function definition) can only be added once at a location.
+We saw this in Listing \ref{example-out} for the decimal-to-string conversion.
+The API will be discussed further in the appendix.
 %When encountering a compound term (e.g.
@@
 
-\subsection{Compositional Reasoning}\label{safety}
-Every statement and expression in an Ace function is governed by exactly one active type, determined by the dispatch protocol, or by the single active base associated with the Ace function.
+\section{Theoretical Foundations}\label{theory}
+\newcommand{\F}[1]{\textsf{#1}~}
+\newcommand{\FF}[1]{\textsf{#1}}
+\newcommand{\Q}{\FF{Arg}}
+\renewcommand{\tnil}[1]{[]}
+\begin{figure}[t]
+\small
+$$\begin{array}{rcl}	
+%\textbf{programs} & 
+\rho & ::= & \F{tycon}\fvar{tycon}~\F{of}\kappaidx~{\{}TC{\}}; \rho \pipe e\\ %\pfam{\familyDf}{\progsort}
+%\\&  \pipe & \pdef{t}{\kappa}{\tau}{\progsort}  \pipe e
+%\pipe \pdef{t}{\kappa}{\tau}{\progsort} 
+ TC & ::= & \F{iana} \{\tau\}; \F{isyn} \{\tau\}; \F{esyn} \{\tau\}; \F{rep} \{\tau\}\\
+\\ 
+e & ::= & \evar{x} \pipe \elam{\evar{x}}{\tau}{e} \pipe I \pipe I : \tau \pipe I :: \fvar{tycon} \pipe E\\
+I & ::= &  \FF{intro}[\tauidx](\splat{e}{1}{n})\\
+E & ::= & e\cdot\FF{elim}[\tauidx](\splat{e}{1}{n})\\
+\\
+ \tau 	& ::= 	& 	\tvar{t} \pipe 
+														\tlam{t}{\kappa}{\tau} \pipe 
+														\tapp{\tau_1}{\tau_2}  
+\\&\pipe&											
+														\tnil{\kappa} \pipe \tcons{\tau_1}{\tau_2} \pipe 
+									                     \tfold{\tau_1}{\tau_2}{h}{t}{r}{\tau_3}
+														\\
+ 		& \pipe	& 	  \ell \pipe 
+		\tunit \pipe 
+														\tpair{\tau_{1}}{\tau_{2}} \pipe 
+														\tfst{\tau} \pipe 
+														\tsnd{\tau} \pipe \cdots 
+														\\	
+%     & \pipe & \tinl{\kappa_2}{\tau_1} \pipe \tinr{\kappa_1}{\tau_2} \\
+%     &\pipe& \tsumcase{\tau}{t}{\tau_1}{t}{\tau_2} \\
+&\pipe	& 	\ttype{tycon}{\tauidx} \\&\pipe& \tfamcase{\tau}{tycon}{x}{\tau_1}{\tau_2}\\
+&\pipe & 					\tifeq{\tau_{1}}{\tau_{2}}{\kappa}{\tau_{3}}{\tau_{4}} 
+%														\\													%				& & \pipe & \tfamcase{\tau}{Fam}{x}{\tau_1}{\tau_2}\\
+%																								
+\\ 		 & 	\pipe	&	\tden{\tau_2}{\tau_1} \pipe \terr %\pipe \tden{\ibar}{\tau}^{\checkmark} 
+\pipe \ttypeof{\tau} \pipe \itransof{\tau} \\% \tdencase{\tau}{x}{t}{\tau_1}{\tau_2}\\
+% %& & \pipe & 
+%%														\tdencase{\tau}{y}{x}{\tau_1}{\tau_2}
+%%														 \\
+%
+ & \pipe & \titerm{\iota}
+			\pipe	\titype{\sigma} \pipe \trepof{\tau} \\
+		& \pipe &  (\Gamma; e)? \pipe \FF{syn}(\tau) \pipe \FF{ana}(\tau; \tau')\\
+%\textbf{external terms} 				&	e	&	::=	&	\evar{x} \pipe 
+%%														\efix{x}{\tau}{e} \pipe 
+%														\elam{\evar{x}}{\tau}{e} \pipe 
+%														\eop{tycon}{op}{
+%															\tauidx
+%														}{
+%  												    		\splat{e}{1}{n}
+%														} \\
+%									& 		&		& 	\\
+%
+%
+%												
+%%\text{deabstracted}& \iota & ::= & \mathcal{G}[\iota, \sigma]\\
+%												
+%\\
+%\\
+%							
+%% &  & \pipe & \tvalof{\tau_1}{\tau_2} \pipe \iup{\tau} \\
+%% 												\trepof{\tau} \pipe \dup{\tau}\\
+%\text{translational IL}	& \bar{\iota} & ::= & x \pipe \ifix{x}{\bar \sigma}{\bar \iota} 
+%	%\pipe \ilam{x}{\bar \sigma}{\bar \iota} \pipe \iapp{\bar \iota_1}{\bar \iota_2} 
+%	\pipe \cdots \pipe \itransof{\tau} \\						
+% & \bar{\sigma} & ::= & \darrow{\bar \sigma_1}{\bar \sigma_2} \pipe \cdots \pipe \dup{\tau} \pipe \trepof{\tau} \\
+%%\text{abstracted} & \sabs & ::= & \darrow{\sabs_1}{\sabs_2} \pipe \cdots \pipe \sabsrep{\tau}
+%											\\
+\kappa	&	::=	&	\karrow{\kappa_1}{\kappa_2} \pipe \klist{\kappa} \pipe \FF{L} \pipe \kunit \pipe \kpair{\kappa_1}{\kappa_2} \pipe \cdots \\
+ & \pipe & \kTypeBlur \pipe \FF{TT} \pipe \kITerm \pipe \kIType \pipe \Q
+%											    \klabel \pipe
+%											    \klist{\kappa} \pipe
+%												\kunit \pipe 
+%												\kpair{\kappa_{1}}{\kappa_{2}} \\
+%												&&\pipe&
+%												\ksum{\kappa_1}{\kappa_2} \pipe
+%												\kTypeBlur \pipe \kDen \pipe 
+%												\kIType								
+%%\textbf{ops signature}			& \Theta	&	::=	&	\kOpEmpty \pipe \kOp{\Theta}{op}{\kappai}\\
+%%											 							&		&		&	\\
+\\
+\\
+\iota	&	::=	&	\evar{x} \pipe 
+												\ifix{\evar{x}}{\sigma}{\iota} \pipe
+												\ilam{\evar{x}}{\sigma}{\iota} \pipe 
+												\iapp{\iota_{1}}{\iota_{2}} 
+												\\ & \pipe & \iunit \pipe \ipair{\iota_1}{\iota_2} \pipe \ifst{\iota} \pipe \isnd{\iota} \pipe \cdots \pipe \iup{\tau}
+%\\&\pipe&								\cdots \pipe 	
+%												n \pipe \iop{\iota_{1}}{\iota_{2}} \pipe \iIfEq{\iota_{1}}{\iota_{2}}{\dint}{\iota_{3}}{\iota_{4}} \\
+%& \pipe & 
+%												\iunit \pipe
+%												\ipair{\iota_{1}}{\iota_{2}} \pipe 
+%												\ifst{\iota} \pipe
+%												\isnd{\iota} 
+%\\&\pipe&												
+%												 \iinl{\sigma_2}{\iota_1} \pipe \iinr{\sigma_1}{\iota_2} \\&\pipe & \icase{e}{x}{e_1}{x}{e_2} \\
+\\	\sigma	&	::=	&    \darrow{\sigma_1}{\sigma_2} \pipe \dunit \pipe \dpair{\sigma_1}{\sigma_2} %\pipe \dint \pipe \dunit \pipe \dpair{\sigma_1}{\sigma_2} \pipe \dsum{\sigma_1}{\sigma_2}  
+\pipe \cdots \pipe \dup{\tau}
+\end{array}$$
+%\vspace{-10pt}
+\caption{\small Syntax of Core $\lamAce$.
+Here, $x$ ranges over external and internal language variables, $\tvar{t}$ ranges over type-level variables, $\fvar{tycon}$ ranges over type constructor names and $\ell$ ranges over labels.
+\label{grammar}}
+\end{figure}
+\begin{figure}
+\small
+\begin{flalign}
+& \F{tycon}\fvar{record}~\F{of}\klist{\kpair{\FF{L}}{\kTypeBlur}}~\{\\
+& \quad \FF{iana}~\{\tlam{i}{\kpair{\klist{\kpair{\FF{L}}{\kTypeBlur}}}{\klist{\FF{L}}}}{\tlam{a}{\klist{\Q}}{\\& \quad\quad\tfold{\tvar{zip3}~\tfst{\tvar{i}}~\tsnd{\tvar{i}}~\tvar{a}}
+	 {\titerm{\tunit}}{h}{t}{r}{
+	\\&\quad\quad\quad\tifeq{\tfst{\tvar{first}~\tvar{h}}}{\tvar{second}~\tvar{h}}{\FF{L}}{
+	\\&\quad\quad\quad\quad
+	\F{let}~\tvar{x}:\kITerm~=~\FF{ana}(\tvar{third}~\tvar{h}; \tsnd{\tvar{first}~\tvar{h}})
+~\FF{in}
+\\ & \quad\quad\quad\quad \tfold{\tvar{t}}{\tvar{x}
+}{\_}{\_}{\_}{\titerm{(\iup{\tvar{x}},\iup{\tvar{r}})}}	}{\terr}
+%     \F{let}~\tvar{tt}:\FF{TT}=~\FF{syn}(\tsnd{\tsnd{\tvar{h}}}) \F{in}\\
+%\\     & \quad\quad\quad 
+}
+}}\}\\
+& \quad \FF{isyn}~\{\tlam{i}{\klist{\FF{L}}}{\tlam{a}{\klist{\Q}}{
+\\&\quad\quad \tfold{\tvar{zip2}~\tvar{i}~\tvar{a}}{\tden{\titerm{()}}{\ttype{record}{\tnil{{\kpair{\FF{L}}{\kTypeBlur}}}}}}{h}{t}{r}{
+\\&\quad\quad\quad \F{let}~\tvar{htt}:\FF{TT}~=~\FF{syn}(\tsnd{\tvar{h}})
+~\FF{in}
+\\&\quad\quad\quad \F{let}~\tvar{hty}:\kTypeBlur~=~\ttypeof{\tvar{htt}}
+\\&\quad\quad\quad \F{let}~\tvar{ty}:\kTypeBlur~=~\ttype{record}{(\tfst{\tvar{h}}, \tvar{hty})::\tnil{{\kpair{\FF{L}}{\kTypeBlur}}}}
+~\FF{in}
+\\&\quad\quad\quad \tfold{\tvar t}{\tden{\itransof{\tvar{htt}}}{\tvar{ty}}}{\_}{\_}{\_}{
+\\&\quad\quad\quad\quad \tden{\titerm{(\iup{\itransof{\tvar{htt}}},\iup{\itransof{\tvar{r}}})}}{\tvar{ty}}}}
+}}\}\\
+& \quad \FF{esyn}~\{\tlam{i}{\FF{L}}{\tlam{a}{\klist{\Q}}{
+	\tvar{arity1}~\tvar{a}~\tlam{ty}{\kTypeBlur}{\tlam{x}{\kITerm}{
+\\&\quad\quad		\tfamcase{\tvar{ty}}{record}{sig}{
+\\&\quad\quad\quad \tfold{\tvar{sig}}{\terr}{h}{t}{r}{
+\\&\quad\quad\quad\quad \tifeq{\tfst{\tvar{h}}}{\tvar{i}}{\FF{L}}{\tfold{\tvar{t}}{\tden{\tvar{x}}{\tsnd{\tvar{h}}}}{\_}{t'}{\_}{
+\\&\quad\quad\quad\quad\quad \tfold{\tvar{t'}}{\tden{\titerm{\ifst{\iup{\tvar{x}}}}}{\tsnd{\tvar{h}}}}{\_}{\_}{\tvar{r'}}{
+\\&\quad\quad\quad\quad\quad\quad \tden{\titerm{\isnd{\iup{{\itransof{\tvar r}}}}}}{\tsnd{\tvar{h}}}}}\\&\quad\quad\quad\quad\hspace{-3px}}{\tvar{r}}
+}\\&\quad\quad\hspace{-3px}}{\terr}
+	}}
+}}\\
+& \quad \FF{rep}~\{\tlam{i}{\klist{\kpair{\FF{L}}{\kTypeBlur}}}{ 	\\& \quad \quad \tfold{\tvar{i}}{\titype{\dunit}}{s}{j}{r}{
+ 		\tfold{\tvar{j}}{\trepof{\tsnd{\tvar{s}}}}{\_}{\_}{\_}{\\
+		&\quad\quad\quad
+ 		\titype{\dpair{\dup{\trepof{\tsnd{\tvar{s}}}}}{\dup{\tvar{r}}}}
+ 		}
+ 	}}\}\\
+& \}; \\
+& \F{let}\tvar{R} : \kTypeBlur = \ttype{record}{(\ell_1, \ttype{record}{\tnil{{\FF{L}\times\kTypeBlur}}}) :: \tnil{{\FF{L}\times\kTypeBlur}}}~\F{in}\\
+&\F{let}id = \elam{x}{\tvar{R}}{\{\ell_1=~x\cdot\ell_1\} :: \fvar{record}}\\
+&\F{let}triv = id[\{\ell_1 = \{\}\}]\cdot\ell_1
+%&\F{let}one = \FF{intro}[\ell_\text{succ}]
+%&\F{let}plus = \elam{x}{\tvar{nat}}{\elam{y}{\tvar{nat}}{\FF{elim}[()](x; y; \\
+%& \quad \elam{r}{\tvar{nat}}{intro[\ell_\text{succ}](r) : \tvar{nat}}})}
+\end{flalign}
+\caption{The definition of the record type in $\lamAce$ using the desugarings in Figure \ref{desugaring} and with the addition of simple let bindings for both type and term variables (not shown).
+Some type-level helper functions also omitted for concision.}
+
+\label{record-theory}
+\end{figure}
+We will now give a core typed lambda calculus, $\lamAce$, that captures the semantics described in the previous sections.
+It is intended to make precise how active typechecking and translation works and how our mechanism relates to existing work on bidirectional typechecking, type-level computation and typed compilation while abstracting away from the details of Python's syntax and imposing a stronger type-level semantics that will allow us to state metatheoretic properties of interest.
+We will assume a fixed base for functions (providing the standard semantics of lambda functions) and target language, which we here call the \emph{internal language}.
+
+The syntax of $\lamAce$ is shown in Figure \ref{grammar} and an example \emph{program} that defines a type constructor, $\fvar{record}$, with a semantics similar to that given in Listing \ref{record}\footnote{Technically, this defines labeled tuples, because the order of labels matters.}, using it to write an identity function and compute the value $triv$ (the empty record), is shown in Figure \ref{record-theory}.
+
+%In this section, we will develop an ``actively typed'' version of the simply-typed lambda calculus with simply-kinded type-level computation called $\lamAce$.
+More specifically, the level of types, $\tau$, will itself form a simply-typed lambda calculus.
+\emph{Kinds} classify type-level terms in the same way that types conventionally classify expressions.
+Types become just one kind  of type-level value (which we will write $\kTypeBlur$, though it is also variously written $\star$, \verb|T| and \verb|Type| in various settings).
+Rather than there being a fixed set of type and operator constructors, we allow the programmer to declare new  constructors, and give their static and dynamic semantics by writing type-level functions.
+The kind system combined with techniques borrowed from the typed compilation literature and a form of type abstraction will allow us to prove strong type safety, decidability and conservativity theorems.
+%
+%The syntax of Core $\lamAce$ is given in Fig.
+\ref{grammar}.
+An example of a program defining type and operator constructors that can be used to construct an active embedding of G\"odel's \textbf{T} into $\lamAce$ is given in Fig.
+\ref{nat}.
+We will discuss its semantics and how precisely the embedding, seen being used starting on line 15 to ultimately compute the sum of two and two, works as we go on.
+Natural numbers can, of course, be isomorphically embedded in existing languages, with a similar usage and asymptotic performance profile (up to function call overhead as an abstract type, for example).
+We will provide more sophisticated examples where this is less feasible later on (and note that type abstraction is an orthogonal mechanism).
+%a type constructor declaration, $\fvar{Nat}$, indexed trivially, together with three operator constructors, also all indexed trivially, that implement the standard introductory forms for natural numbers as well as the recursor operator (as in G\"odel's T \cite{pfpl}).
+Following the type constructor declaration, we apply $\fvar{Nat}$ with the trivial index, $\tunit$, to form the type $\tvar{nat}$.
+Finally, we write an external term that uses the operators associated with $\fvar{nat}$ and the built-in constructor $\fvar{Parr}$, governing partial functions, to define an addition function and compute the addition of the natural numbers  two and two.
+We will introduce a more convenient concrete syntax in later portions of this thesis; for now we will restrict ourselves to the abstract syntax so that this example can directly aid in understanding the semantics.
+
+\subsection{Overview}\label{programs}
+
+A \emph{program}, $\rho$, consists of a series of constructor declarations followed by an external term, $e$.
+Compiling a program consists of first \emph{kind checking} it (see below), then typechecking the external term and simultaneously {translating} it to a term, $\iota$, in the {typed internal language}.
+%These correspond to the premises of the \emph{central compilation judgement} $\pkcompiles{\rho}{\iota}$:
+%\[
+%\inferrule[p-compiles]{
+%	\emptyset \vdash_{\fvalCtx_0} \rho\\
+%%	\progOK{\emptyset}{\fvalCtx_0}{\rho}\\
+%	\pcompiles{\fvalCtx_0}{\rho}{\iota}
+%}{\pkcompiles{\rho}{\iota}}
+%\]
+%This anchors our exposition; we will describe how it is derived (i.e. how to write a compiler for $\lamAce$) in the following sections. 
+The key judgements are the \emph{bidirectional active typing judgements}  (Fig.
+\ref{att}, which we describe starting in Sec.
+\ref{opcons}).
+They relate an external term, $e$, to a {type}, $\tau$, called its \emph{type assignment}, and an internal term, $\iota$, called its \emph{translation}, under \emph{typing context} $\Gamma$ and \emph{constructor context} $\Phi$.
+The first is synthesis, the second analysis.
+\[\Gamma \vdash_\fvalCtx e \Rightarrow \tau \leadsto \iota
+~~~~~~~
+\Gamma \vdash_\fvalCtx e \Leftarrow \tau \leadsto \iota\]
+%\[\ecompilesAX{e}{\tau}{\iota}\]
+
+The typing context, $\Gamma$, maps variables to types in essentially the conventional way (\cite{pfpl} contains the necessary background for this section).
+The constructor context, $\Phi$, tracks user-defined type  constructors.
+% Weakening and exchange of the constructor context closely related to the issue of conservativity that we will return to.
+Each constructor is identified by name, so there is no analog to contraction.
+
+The dynamic behavior of an external term is determined entirely by its translation to the internal language, which has a conventional operational semantics.
+This form of semantics can be seen as lifting into the language specification the first stage of a type-directed compiler like the TIL compiler for Standard ML \cite{tarditi+:til-OLD} and has some parallels to the Harper-Stone semantics for Standard ML, where external terms were also given meaning by elaboration from the EL to an IL \cite{Harper00atype-theoretic}.
+%
+
+In $\lamAce$, the internal language (IL) provides partial functions (via the generic fixpoint operator of Plotkin's PCF) and simple product  types for the sake of our example.
+In practice, the internal language could be any typed  language with a specification for which type safety and decidability of typechecking have been satisfyingly determined.
+The internal type system serves as a ``floor'': guarantees that must hold for terms of any type (e.g.
+that out-of-bounds access to memory never occurs) must be maintained by the internal type system.
+User-defined constructors can enforce invariants stronger than those the internal type system maintains at particular types, however.
+Performance is also ultimately limited by the internal language and downstream compilation stages that we do not here consider (safe compiler extension has been discussed in previous work, e.g.
+\cite{conf/pldi/TatlockL10}).
+
+\subsection{Types and Type-Level Computation}\label{types}
+$\lamAce$ supports, and makes extensive use of, simply-kinded type-level computation.
+Specifically, type-level terms, $\tau$, themselves form a typed lambda calculus.
+The classifiers of type-level terms are called \emph{kinds}, $\kappa$, to distinguish them from  \emph{types}.
+Types are  type-level values of kind $\kTypeBlur$.
+In Ace, the type-level language (together with the level of programs) is written in Python, which can be thought of as having a rather limited kind system (with one kind, \verb|dyn|).
+Here, we are able to more precisely discuss the kinds of values in the type-level language.
+
+In most languages, types are formed by applying one of a collection of \emph{type constructors} to zero or more \emph{indices}.
+In $\lamAce$, the situation is notionally similar.
+User-defined type constructors can be declared at the top of a program (or lifted to the top, in practice) using \textsf{tycon}.
+Each constructor in the program must have a unique name, written e.g.
+\fvar{record}.\footnote{We assume naming conflicts can be avoided by some extrinsic mechanism.} A type constructor must also declare an \emph{index kind}, $\kappaidx$.
+
+ To permit the embedding of interesting type systems, the type-level language includes several kinds other than $\kTypeBlur$.
+We functional data structures to the type level: here, only unit ($\kunit$), binary products ($\kpair{\kappa_1}{\kappa_2}$) and lists ($\klist{\kappa}$), in addition to labels (introduced as $\ell$, possibly with a subscript, having kind \FF{L}).
+Our record type constructor is indexed by a list of pairs of {labels}  and types (a signature, in essence; line 1).
+The type constructor $\fvar{arrow}$ is included in the initial constructor context, $\fvalCtx_0$ and has index kind $\kpair{\kTypeBlur}{\kTypeBlur}$.
+ In practice, one could include a richer functional programming language and retain the spirit of the calculus, as long as it does not introduce general recursion at the type level.
+
+A type is introduced by applying a type constructor to an index of this kind, written $\ttype{Tycon}{\tauidx}$.
+For example, the type of natural numbers is indexed trivially, so it is written $\ttype{Nat}{\tunit}$.
+We see a record type, abbreviated $\tvar{R}$, constructed on line 26.
+
+The kind $\kTypeBlur$ also has an elimination form, $$\tfamcase{\tau}{Tycon}{x}{\tau_1}{\tau_2}$$ allowing the extraction of a type index by case analysis against a contextually-available type constructor.
+To a first approximation, one might think of type constructors as constructors of a built-in open \cite{conf/ppdp/LohH06}, $\kTypeBlur$, at the type-level.
+Like open datatypes, there is no notion of exhaustiveness so the default case is required for totality.
+We will see where this is used shortly.
+
+%We will write the kind of types as $\kTypeBlur$, though it is also written $\star$ or \verb|Type| in various similarly structured languages (see Sec. \ref{related-work}). 
+% Rather than there being a fixed set of type constructors, we allow the programmer to declare new type  constructors, and give the static and dynamic semantics of their associated operators, by writing type-level functions.
+In the semantics for this calculus, our kind system combined with techniques borrowed from the typed compilation literature and a form of type abstraction allow us to prove strong type safety, decidability and conservativity theorems.
+
+% and integers ($\dint$).
+We also include labels ($\klabel$), written in a slanted font, e.g.
+$\tlabel{myLabel}$, which are string-like values that only support comparison and play a distinguished role in the expanded syntax, as we will later discuss.
+Our first example, $\fvar{nat}$, is indexed trivially, i.e.
+by unit kind, $\kunit$, so there is only one natural number type, $\ttype{nat}{\tunit}$, but we will show examples of type constructors that are indexed in more interesting ways in later portions of this work.
+For example, $\fvar{LabeledTuple}$ has index kind $\klist{\kpair{\klabel}{\kTypeBlur}}$.
+
+ Type constructors are not first-class; they do not themselves have arrow kind as in some kind systems (e.g.
+\cite{watkins2008specifying}; Ch.
+22 of \emph{PFPL} describes a related system \cite{pfpl}).
+The type-level language does, however, include total functions of arrow kind, written $\karrow{\kappa_1}{\kappa_2}$.
+Type constructor application can be wrapped in a type-level function to emulate a first-class or uncurried version of a type constructor for convenience (indeed, such a wrapper could be generated automatically, though we do not do so).
+
+Two type-level terms of kind $\kTypeBlur$ are equivalent if they apply the same constructor, identified by name, to equivalent indices.
+Going further, we ensure that deciding type equivalence requires only checking for syntactic equality after normalization by imposing the restriction that equivalence at a type constructor's index kind must be decidable in this way.
+Our treatment of equivalence in the type-level language is thus quite similar to the treatment of term-level equality using ``equality types'' in a language like Standard ML.
+% A kind $\kappa$ is an  \emph{equality kind} if $\kEq{\kappa}$ can be derived (see appendix). 
+Conditional branching on the basis of equality at an equality kind can be performed in the type-level language.
+Equivalence at arrow kind is not decidable by our criteria, so type-level functions cannot appear within type indices.
+This also prevents general recursion from arising at the type level.
+Without this restriction, a type-level function taking a type as an argument could ``smuggle in'' a self reference as a type index, extracting it via case analysis (continuing our analogy to open datatypes, this is closely related to the positivity condition for inductive datatypes in total functional languages like Coq).% as maintaining the metatheoretic guarantee that typing respects type equivalence would impose a substantial burden in such a setting.% (a na\"ive approach to this would impose non-trivial extrinsic proof obligations onto extension developers that, unlike in others in this thesis, could threaten type safety).
+
+Every type constructor also defines a \emph{representation}, a type-level function that associates with every type an internal \emph{representation type} (analagous to \verb|trans_type| above).
+We will return to this after introducing the external forms and operator definitions.
+
+\subsection{Core External Forms and Desugaring}\label{opcons}
+The syntax for external terms (Figure \ref{grammar}) contains variables, $\lambda$ terms, three generalized introductory forms and a single generalized elimination form.
+The introductory forms are either unascribed, ascribed with a type or ascribed with a type constructor, as in our discussion of Ace.
+These generalized forms take a single a type-level value as an {index} and $n \geq 0$ arguments, which are other external terms.
+To better motivate this choice, we can give a purely syntactic desugaring of a Python-like syntax with labels to these forms, shown in Figure \ref{desugaring}.
+It is instructive to rewrite lines 27-28 of Figure \ref{record-theory} using these desugarings.% In Ace, desugarings can be user-defined
+\begin{figure}[t]
+\small
+\[
+\begin{array}{rcl}
+\{ \ell_1{=}~e_1, \ldots, \ell_n{=}~e_n \} & := & \FF{intro}[\ell_1 :: \ldots :: \ell_n :: []](e_1; \ldots; e_n)\\
+(e_1, ..., e_n) & := & \FF{intro}[()](e_1; \ldots; e_n)\\
+\#n & := & \FF{intro}[n]()\\
+"s" & := & \FF{intro}[s]()\\
+e.\ell & := & e\cdot\FF{elim}[\ell]()\\
+e[e_1; \ldots; e_n] & := & e\cdot\FF{elim}[()](e_1; \ldots; e_n)\\
+e.\ell(e_1; ...; e_n) & := & e\cdot\FF{elim}[\ell](e_1; \ldots; e_n)
+%(e_1, ..., e_n) & := & intro[()](e_1; ...; e_n)
+%n &  := & intro[n]()
+%s & := & intro[s]()
+%
+%e.l & := & elim[l](e)
+%e[e_1; ...; e_n] & := & elim[()](e_1, ..., e_n)
+%e.l(e_1; ...; e_n) & := & elim[l]
+\end{array}
+\]
+\caption{Desugaring from conventional syntax to core forms. We assume that the type-level language has numbers, $n$, and strings, $s$ (not shown for concision).}
+\label{desugaring}
+\end{figure}
+
+%User-defined operator constructors are declared using \textsf{opcon}.
+For reasons that we will discuss, our calculus associates every operator  constructor with a type constructor.
+The \emph{fully-qualified name} of every operator constructor, e.g.
+$\fvar{Nat}.\opvar{z}$, must be unique.
+Operator constructors, like type constructors, declare an index kind, $\kappaidx$.
+In our first example, all the operator constructors are indexed trivially (by index kind $\kunit$), but other examples use more interesting indices.
+For example, in SML, the projection operator \verb|#3| can be applied to an $n$-tuple, $e$, iff $n \geq 3$.
+Note that it thus cannot be a function with a standard arrow type.
+Notionally, \verb|#| is an operator constructor and \verb|3| is its index.
+In an active embedding of $n$-tuples into $\lamAce$, this would be written $\eop{Tuple}{prj}{3}{e}$ (we will nearly recover ML's syntax later).
+$\fvar{LabeledTuple}.\opvar{prj}$ is the operator constructor used to access a field of a labeled tuple, so it has index kind $\klabel$.
+An operator itself is, notionally, selected by indexing an operator constructor, e.g.
+$\fvar{Nat}.\opvar{s}\langle \tunit \rangle$, but technically neither operator constructors nor operators are first-class at any level (additional machinery would be needed, e.g.
+an \textsf{Op} kind, but this is not fundamental to our calculus).
+%Instead, in the external language, an operator constructor is applied by simultaneously providing an index and  $n \geq 0$ \emph{arguments}, written $\eop{Tycon}{op}{\tauidx}{\splat{e}{1}{n}}$\footnote{It may be helpful to distinguish between type/operator constructors and \emph{term formers}.
+There are term formers at all levels in the calculus.
+For example, operator constructor application and $\lambda$ are  external term formers, and type constructor application is a type-level term former.
+We might write these following Harper's conventions for abstract syntax to highlight this distinction \cite{pfpl}: $\mathtt{lam}[\tau](x.e)$, $\mathtt{ocapp}[\fvar{Tycon}, \opvar{op}, \tauidx](\splat{e}{1}{n})$ and $\texttt{tcapp}[\fvar{Tycon}](\tau)$.}.
+For example, on line 18 of Fig.
+\ref{nat}, we see the operator constructors $\fvar{Nat}.\opvar{z}$ and $\fvar{Nat}.\opvar{s}$ being applied to compute $two$.
+%\footnote{Although our focus here is entirely on semantics, a brief note on syntax: in the expanded syntax, the trivial indices and empty argument lists can be omitted, so we could write \texttt{Nat.s(Nat.s(Nat.z))}.
+With the ability to ``open'' a type's operators into the context, we could shorten this still to \texttt{s(s(z))}.
+Alternatively, with the ability to define a TSL in a manner similar to that in Sec.
+\ref{aparsing}, we might instead just write \texttt{2}.}
+
+\subsection{Operator Definitions and Representational Consistency}
+
+%
+\newcommand{\atjsynX}[3]{\Gamma \vdash_\fvalCtx #1 \Rightarrow #2 \leadsto #3}
+\newcommand{\atjanaX}[3]{\Gamma \vdash_\fvalCtx #1 \Leftarrow #2 \leadsto #3}
+\begin{figure}[t]
+\small
+$\fbox{\inferrule{}{\ecompilesAX{e}{\tau}{\iota}}}$
+~~~~$\Gamma ::= \emptyset \pipe \Gamma, x \Rightarrow \tau$
+\begin{mathpar}
+\inferrule[att-var]{
+	x \Rightarrow \tau \in \Gamma
+}{
+	\atjsynX{x}{\tau}{x}
+}
+
+\inferrule[att-syn-to-ana]{
+	\atjsynX{e}{\tau}{\iota}
+}{
+	\atjanaX{e}{\tau}{\iota}
+}
+
+\inferrule[att-lam]{
+	\tau_1 \Downarrow_\fvalCtx \tau_1'\\
+		\trepof{\tau_1'} \Downarrow_\fvalCtx \titype{\sigma}\\\\
+	\Gamma, x \Rightarrow \tau_1' \vdash_\fvalCtx e \Rightarrow \tau_2 \leadsto \iota
+%	\iota \hookrightarrow_\fvalCtx \iota'\\
+%		\sigma \hookrightarrow_\fvalCtx \sigma'
+%	\ddbar{\fvar{Arrow}}{\fvalCtx}{\trepof{\tau_1'}}{\sbar_1}\\
+	%\delfromtau{$\Xi_0$}{\fvalCtx}{\tau_1'}{\sabs}\\\\
+}{
+	\atjsynX{\elam{x}{\tau_1}{e}}{\ttype{arrow}{(\tau_1', \tau_2)}}{\ilam{x}{\sigma'}{\iota}}
+}
+
+\inferrule[att-i-unasc]{
+	\vdash_\fvalCtx \FF{iana}(\fvar{tycon})=\taudef\\
+	\taudef~(\tauidx, \tauidx')~((\Gamma; e_1)? :: \ldots :: (\Gamma; e_n)? :: []) \Downarrow_\fvalCtx \titerm{\iota}\\
+%	\trepof{\ttype{tycon}{\tauidx'}} \Downarrow_\fvalCtx \titype{\sigma}\\
+		\trepof{\ttype{tycon}{\tauidx'}} \Downarrow_\fvalCtx \titype{\sigma}\\
+		\Gamma \vdash_\fvalCtx \iota : \sigma
+}{
+	\atjanaX{\FF{intro}[\tauidx](e_1; \ldots; e_n)}{\ttype{tycon}{\tauidx'}}{\iota}
+}
+
+\inferrule[att-i-asc-ty]{
+	\atjanaX{I}{\tau}{\iota}
+}{
+	\atjsynX{I : \tau}{\tau}{\iota}
+}
+
+\inferrule[att-i-asc-tycon]{
+	\vdash_\fvalCtx \FF{isyn}(\fvar{tycon})=\taudef\\\\
+	\taudef~\tauidx~(\Gamma; e_1)? {::}{\ldots}{::}(\Gamma; e_n)? {::} []) \Downarrow_\fvalCtx \tden{\titerm{\iota}}{\ttype{tycon}{\tauidx'}}\\
+			\trepof{\tau_1'} \Downarrow_\fvalCtx \titype{\sigma}\\
+	\Gamma \vdash_\fvalCtx \iota : \sigma
+}{
+	\atjsynX{\FF{intro}[\tauidx](e_1; \ldots; e_n)] :: \fvar{tycon}}{\tau}{\iota'}
+}
+
+\inferrule[att-elim]{
+	\atjsynX{e}{\ttype{tycon}{-}}{-}\\
+	\vdash_\fvalCtx \FF{esyn}(\fvar{tycon})=\taudef\\
+	\taudef~((\Gamma; e)? :: (\Gamma; e_1)? :: \ldots :: (\Gamma; e_n)? :: []) \Downarrow_\fvalCtx \tden{\tau}{\iota}\\
+			\trepof{\tau} \Downarrow_\fvalCtx \titype{\sigma}\\
+	\Gamma \vdash_\fvalCtx \iota : \sigma
+}{
+	\atjsynX{e\cdot\FF{elim}[\tauidx](e_1; \ldots; e_n)}{\tau}{\iota}
+}
+\end{mathpar}
+\caption{\small The active typing judgement. The normalization judgement for type-level terms ($\Downarrow$) and the representational consistency check will be in an appendix.}
+\label{atj}
+\end{figure}
+\begin{figure}
+\small
+\begin{mathpar}
+\inferrule[repof]{
+	\tau \Downarrow_\fvalCtx \ttype{tycon}{\tauidx}\\
+	\vdash_\fvalCtx \FF{rep}(\fvar{tycon}) = \taurep\\
+	\taurep \tauidx \Downarrow_\fvalCtx \titype{\sigma}
+}{
+	\FF{repof}(\tau) \Downarrow_\fvalCtx \titype{\sigma}
+}
+
+\inferrule[syn]{
+    \tau \Downarrow_\fvalCtx (\Gamma; e)?\\
+	\atjsynX{e}{\tau}{\iota}
+}{
+	\FF{syn}(\tau) \Downarrow_\fvalCtx \tden{\titerm{\iota}}{\tau}
+}
+
+\inferrule[ana]{
+	\tau_1 \Downarrow_\fvalCtx (\Gamma; e)?\\
+	\tau_2 \Downarrow_\fvalCtx \ttype{tycon}{\tauidx}\\\\
+	\atjanaX{e}{\ttype{tycon}{\tauidx}}{\iota}\\
+}{
+	\FF{ana}(\tau_1; \tau_2) \Downarrow_\fvalCtx \titerm{\iota}
+}
+\end{mathpar}
+\caption{\small Evaluation semantics for the type-level language.
+Missing rules (including error propagation rules, which immediately cause failure of typechecking) are unsurprising and will be given in an appendix.}
+\label{tleval}
+\end{figure}
+The expressive and metatheoretic power of the calculus arises from how the rules for the active typing judgement handle these  generalized forms.
+Rather than fixing the specification of a finite collection of operator constructors and tasking the \emph{compiler} with deciding a typing derivation on its basis, the specification instead delegates to a type-level function associated with a type constructor.
+In the core calculus, this is one of three functions, called \FF{iana}, \FF{isyn} and \FF{esyn} in the grammar.\footnote{This means that any particular type constructor only supports a single introduction and elimination desugaring.
+This is a minor inconvenience in some cases that is resolved in Ace by the use of methods.} We see how this is done with the rules for the active typing judgements, given in Figure \ref{atj}.
+
+The first two rules are standard in bidirectional type systems: variables synthesize types and if a term synthesizes a type, it can be analyzed against a type (cf.
+Listing \ref{context}).
+Variables translate to variables, and if we are simply converting from synthesis to analysis, translation is not affected.
+
+The rule $\fvar{att-lam}$ must take into account the fact that, because we support type-level computation, the type annotation on the argument may not be in normal form.
+Thus, we evaluate it to normal form.
+Note that the kinding rules (not shown) will guarantee that, because $\tau_1$ is of kind $\kTypeBlur$, its normal form, $\tau'_1$, is of the form $\ttype{tycon}{\tauidx}$ for some normal $\tauidx$.
+To generate an appropriate internal type annotation in the translation, we need to compute the representation type of $\tau_1'$.
+This involves calling the representation function associated with its type constructor (rule \fvar{repof} in Figure \ref{tleval}).
+The form $\titerm{\sigma}$ is a \emph{quoted internal type}.
+Note in the syntax that there is an unquote form, $\dup{\tau}$, which allows us to compose internal types compositionally without needing to expose an elimination form.
+Indeed, it is an interesting facet of our calculus that we never need to examine syntax trees directly to implement extensions.
+The evaluation semantics remove quotations, so the normal form of a quoted internal type contains an internal type with no quotations.
+
+Lambda functions are the only introductory form requiring special support in the calculus (because they need to  manipulate the context; see Discussion).
+The next three rules show how any other abstractions that we define (e.g.
+records, decimals, etc.) make use of a generalized introductory form.
+
+The rule  $\fvar{att-i-unasc}$ shows that unascribed introductory forms can only be analyzed against a type.
+Given such a type, the rule extracts a definition, named $\FF{iana}$, from the type constructor (cf.
+the \verb|ana_Dict| and \verb|trans_Dict| methods earlier).
+It calls this function with a pair containing the operator and type index and a list of \emph{reified arguments}, which have kind $\Q$ and introductory form $(\Gamma; e)?$.
+These are only constructed by the compiler (there would be no corresponding form in the concrete syntax).
+Their purpose is to allow the definition to programatically invoke synthesis and analysis as needed using the $\FF{ana}$ and $\FF{syn}$ operators.
+We can see the $\fvar{record}$ type constructor doing so on line 5 of Figure \ref{record-theory} to ensure that the field value provided as an argument has the same type as the corresponding label (accomplished by simultaneously folding over all three pieces of input data).
+The rule for performing analysis, $\fvar{ana}$, is in Figure \ref{tleval}.
+If it succeeds, analysis returns a \emph{translation}, which is a quoted internal term with kind $\kITerm$ and introductory form $\titerm{\iota}$.
+Like quoted internal types, there is an unquote form that is eliminated during evaluation.
+The operator definition uses this to construct a translation for the record.
+As before, empty records translate to units and records with a single field are unadorned.
+In this example, records with two or more fields translate to nested tuples.
+
+Because we have a representation type associated with the type, we can check that the translation is \emph{representationally consistent} (the final premise).
+As we will discuss, representational consistency combined with type safety of the internal language implies type safety overall (it arises as a strengthening of the inductive hypothesis needed to prove type safety, and is closely related to work on \emph{type-preserving compilation} in the TIL compiler for Standard ML, \cite{tarditi+:til-OLD}).
+
+The next rule, $\fvar{att-i-asc-ty}$ states that introductory forms ascribed with a type are analyzed against that type.
+
+Introductory forms ascribed with a type constructor can, according to the final introductory rule, $\fvar{att-i-asc-tycon}$, synthesize a type via the definition $\FF{isyn}$.
+It is passed the operator index (there is no type index, since we only have a type constructor) and a list of arguments, as before.
+In this case, it must return not just a translation, but also a type.
+We use the form $\tden{\tau_2}{\tau_1}$ for such a pairing, which has kind $\FF{TT}$.
+The type and translation can be extracted from it using the appropriately named elimination forms (indeed, as given, it is merely a pair, but we give it special syntax for clarity -- it looks like the conclusion of the typing judgement -- and for other reasons that will become clear in future work).
+To synthesize a type from a list of labels and arguments, we must be able to synthesize types from the arguments.
+The $\FF{syn}(\tau)$ operator permits this, per rule $\fvar{syn}$, seen being used in Listing \ref{record-theory}.
+We check representational consistency of the result, as before.
+
+Finally, we show the rule for elimination forms.
+It operates by first synthesizing a type for the ``primary'' subterm, then extracting the $\FF{esyn}$ definition from its type constructor.
+As before, it is called with the arguments and representational consistency is checked.
+Note that the primary subterm is itself the first argument (though we have already synthesized a type for it, it is more uniform and clear to allow the operator  to do so again, which is done in our example by the $\tvar{arity1}$ helper function).
+
+\subsection{Metatheory}
+The metatheoretic properties of interest are: type safety and termination of the type-level language (guaranteed by our kind system, though the details will need to be provided in an appendix; termination is non-trivial), type safety and decidability of the internal language (it is a standard variant of PCF, so this is trivial) and representational consistency (discussed above, follows because we check it explicitly).
+Type safety for the language as a whole comes as a corollary of these lemmas.
+We plan to provide detailed proofs of these, but for now, they should be treated as conjectures and our formulation above as expository.
+We believe that despite this, the formalization is surprisingly elegant and concise, given its expressive power.
+It is a useful exercise to implement natural numbers, ala G\"odel's \textbf{T}, using this calculus, assuming that the type-level and internal languages include integers and one can lift them from the former into the latter.
+%%
+%
+%For the synthetic functions, it must decide the type assignment and the translation on the basis of the index and arguments, or decide that this is not possible.
+An operator constructor with index kind $\kappaidx$ must have a definition of kind \[\kappaidx \rightarrow \klist{\kDen} \rightarrow (\kDen + \kunit)\]
+%As this suggests, the active typing judgement  will provide the operator index and a list of {recursively determined \emph{derivates}}, which have kind $\kDen$, for the arguments and ask the definition to return a value of ``option kind'', $\ksum{\kDen}{\kunit}$, where the trivial case indicates that a derivate cannot be constructed due to an invalid index, an incorrect number of arguments or an argument of an invalid type.\footnote{In practice, we would require operator constructor providers to report information about the precise location of the error (e.g.
+which argument was of incorrect type) and provide an appropriate error message and other metadata, but we omit this in the semantics.}
+%
+%
+%A \emph{program}, $\rho$, consists of a series of type constructor declarations followed by an external term, $e$.
+The syntax for external terms contains six forms: variables, $\lambda$ terms, a generalized unascribed introductory form, a generalized introductory form ascribed with a type, a generalized introductory form ascribed with a type constructor and a generalized elimination form.
+We will discuss how a more natural syntax  like Python's can be added by simple desugaring.\todo{syntax}
+%
+%
+%
+%Compiling a program consists of first \emph{kind checking} it (typechecking the type-level terms, which has no analog in Ace because the type-level language is Python; discussed further below), then typechecking the external term and {translating} it to a term, $\iota$, in the {typed internal language}.
+%These correspond to the premises of the \emph{central compilation judgement} $\pkcompiles{\rho}{\iota}$:
+%%\[
+%%\inferrule[p-compiles]{%
+%%	\emptyset \vdash_{\fvalCtx_0} \rho\\
+%%	\progOK{\emptyset}{\fvalCtx_0}{\rho}\\
+%%	\pcompiles{\fvalCtx_0}{\rho}{\iota}
+%%}{\pkcompiles{\rho}{\iota}}
+%%\]
+%%his anchors our exposition; we will describe how it is derived (i.e. how to write a compiler for $\lamAce$) in the following sections. 
+%
+%The key judgement in the calculus is the \emph{active typing judgement}  (Fig.
+\ref{att}, which we describe starting in Sec.
+\ref{opcons}).
+It relates an external term, $e$, to a {type}, $\tau$, called its \emph{type assignment}, and an internal term, $\iota$, called its \emph{translation}, under \emph{typing context} $\Gamma$ and \emph{constructor context} $\Phi$:
+%\[\ecompilesX{e}{\tau}{\iota}\]
+%
+%The typing context $\Gamma$ maps variables to types in essentially the conventional way (\cite{pfpl} contains the necessary background for this paper).
+The constructor context tracks user-defined type and operator constructors.
+% Weakening and exchange of the constructor context closely related to the issue of conservativity that we will return to.
+Each constructor is identified by name, so there is no analog to contraction.
+%There is no separate operational semantics for the external language.
+Instead, the dynamic behavior of an external term is determined by its translation to the internal language, which has a more conventional operational semantics.
+This form of semantics can be seen as lifting into the language specification the first stage of a type-directed compiler like the TIL compiler for Standard ML \cite{tarditi+:til-OLD} and has some parallels to the Harper-Stone semantics for Standard ML, where external terms were also given meaning by elaboration from the EL to an IL \cite{Harper00atype-theoretic}.
+%
+%
+%In $\lamAce$, the internal language (IL) provides partial functions (via the generic fixpoint operator of Plotkin's PCF), simple product and sum types and a base type of integers (to make our example interesting and as a nod toward speed on contemporary machines).
+In practice, the internal language could be any typed  language with a specification for which type safety and decidability of typechecking have been satisfyingly determined.
+The internal type system serves as a ``floor'': guarantees that must hold for \emph{all} well-typed terms, independent of the constructor context (e.g.
+that out-of-bounds access to memory never occurs), must be maintained by the internal type system.
+User-defined constructors can (only) enforce invariants stronger than those the internal type system maintains.
+Performance is also ultimately limited by the internal language and downstream compilation stages that we do not here consider (safe compiler extension has been discussed in previous work, e.g.
+\cite{conf/pldi/TatlockL10}).
+%
+
+%\begin{figure}
+%\small
+%\begin{flalign}
+%\small
+% & \family{Tuple}{\klist{\kTypeBlur}}{\\
+% & \quad \tops{new}{\kunit}{\_}{a}{\tlam{\_}{\kunit}{\tlam{a}{\klist{\kDen}}{\\
+% & \quad\quad
+% 	\tfold{\tvar{a}}{\tden{{()}}{\ttype{Tuple}{\tnil{\kTypeBlur}}}}{d}{b}{r}{\\
+%			& \quad\quad\quad \tfamcase{\ttypeof{\tvar{r}}}{Ntuple}{i}{\\
+%			& \quad\quad\quad\quad \tfold{\tvar{b}}{\tden{\itransof{\tvar{d}}}{\ttype{Ntuple}{\tcons{\ttypeof{\tvar{d}}}{\tvar{i}}}}\\& \quad\quad\quad\quad}{\_}{\_}{\_}{\tden{\\& \quad\quad\quad\quad\quad {(\itransof{\tvar{d}},\itransof{\tvar{r}})}}{\ttype{Ntuple}{\tcons{{\ttypeof{\tvar{d}}}}{\tvar{i}}}}}
+%			}{\terr}
+%	}
+% }}};\\
+% & \quad \tops{prj}{\kint}{i}{a}{\lambda \tvar{i}:\kint.\lambda \tvar{a}:\klist{\kDen}. 
+% 	\tapp{\tvar{arity1}}{\tapp{\tvar{a}}{\tlam{d}{\kDen}{\\
+%	& \quad\quad\tfamcase{\ttypeof{\tvar{d}}}{Ntuple}{nl}{\\
+% & \quad\quad \tfold{\tvar{nl}}{\terr}{t1}{j}{\_}{\\
+% & \quad\quad \tfold{\tvar{j}}{\tifeq{\tvar{i}}{1}{\dint}{\tden{\itransof{d}}{\tvar{t1}}}{\terr}\\&\quad\quad\quad}{\_}{\_}{\_}{\\
+% & \quad\quad\quad (\tlam{p}{\kpair{\kDen}{\dint}}{\tifeq{\tsnd{\tvar{p}}}{\tvar{i}}{\dint}{\tfst{\tvar{p}}}{\terr}})\\
+% & \quad\quad\quad (\tapp{\tvar{foldl}}{\tapp{\tvar{nl}}{\tapp{(\tden{\tvar{x}}{\tvar{nt}},0)}{
+% 	\tlam{r}{\kpair{\kDen}{\dint}}{\tlam{t}{\kTypeBlur}{\tlam{ts}{\klist{\kTypeBlur}}{\\
+%	& \quad\quad\quad\quad \tifeq{\tvar{i}}{\tsnd{\tvar{r}}}{\dint}{\tvar{r}}{\tdencase{\tfst{\tvar{r}}}{rx}{\_}{\\
+%	& \quad\quad\quad\quad\quad \tifeq{\tvar{i}}{\tsnd{\tvar{r}}+1}{\dint}{\\
+%	& \quad\quad\quad\quad\quad\quad \tifeq{\tvar{ts}}{\tnil{\kTypeBlur}}{\klist{\kTypeBlur}}{(\tden{\tvar{rx}}{\tvar{t}},\tvar{i})\\
+%	& \quad\quad\quad\quad\quad\quad}{(\tden{\titerm{\ifst{\iup{\tvar{rx}}}}}{\tvar{t}},\tvar{i})}\\	
+%	& \quad\quad\quad\quad\quad}{(\tden{\titerm{\isnd{\iup{\tvar{rx}}}}}{\tvar{t}},\tsnd{\tvar{r}}+1)}}{\terr}}}}}
+% }}})
+% }}{\terr}}}}}
+% }\\ & 
+%%& \quad \tops{pr}a}{b}{c}{d}
+% }{d}{i}{(\tlam{i}{\klist{\kTypeBlur}}{
+% 	\\& \quad \quad \tfold{\tvar{i}}{\titype{\dunit}}{s}{j}{r}{
+% 		\tfold{\tvar{j}}{\titype{\trepof{\tvar{s}}}}{\_}{\_}{\_}{\\
+%		&\quad\quad\quad
+% 		\titype{\dpair{\trepof{\tvar{s}}}{\dup{\tvar{r}}}}
+% 		}
+% 	}
+% })}
+%\end{flalign}
+%\caption{ABC}
+%\label{tuple}
+%\end{figure}
+%\begin{figure}[t]
+%\small
+%$\fbox{\inferrule{}{\progOKX{\progsort}}}$
+%~~~$\tvarCtx ::= \emptyctx \pipe \tvarCtxX{t}{\kappa}$
+%~~~$\fvalCtx ::= \emptyctx \pipe \fvalCtxX{\fvalDf}$
+%%$\fCtx ::= \Sigma_0	 \pipe \fvalCtxX$
+%\begin{mathpar}
+%\inferrule[k-tycon]{
+%	\fvar{tycon} \notin \text{dom}(\fvalCtx)\\
+%	\kEq{\kappaidx}\\
+%	\tKind{\tvarCtx}{\fvalCtx}{\taurep}{\karrow{\kappaidx}{\kIType}}\\
+%	\opType{\tvarCtx}{\fvalCtxX{\fvalDf}}{\theta}{\Theta}\\
+%	\progOK{\tvarCtx}{\fvalCtxX{\fvalDf}}{\rho}
+%}{
+%	\progOKX{\pfam{\familyDf}{\rho}}
+%}
+%
+%%\inferrule[def-kinding]{
+%%	\tKindX{\tau}{\kappa}\\
+%%	\progOK{\tvarCtxX{t}{\kappa}}{\fvalCtx}{\rho}
+%%}{
+%%	\progOKX{\pdef{t}{\kappa}{\tau}{\rho}}
+%%}
+%%
+%\inferrule[k-e-prog]{
+%	\exprOK{\tvarCtx}{\fvalCtx}{e}
+%}{
+%	\progOKX{e}
+%}
+%\end{mathpar}
+%$\fbox{$\opType{\tvarCtx}{\fvalCtx}{\theta}{\Theta}$}$
+%\begin{mathpar}
+%\inferrule[k-opcon]{
+%	\tKind{\tvarCtx}{\fvalCtx}{\taudef}{\karrow{\kappaidx}{\karrow{\klist{\kDen}}{(\ksum{\kDen}{\kunit})}}}
+%}{
+%	\opType{\tvarCtx}{\fvalCtx}{{\tops{op}{\kappaidx}{i}{a}{\taudef}}}{
+%	\kOpS{op}{\kappaidx}}
+%}
+%
+%\inferrule[k-opcons]{
+%	\opType{\tvarCtx}{\fvalCtx}{\theta_1}{\Theta_1}\\
+%	\opType{\tvarCtx}{\fvalCtx}{\theta_2}{\Theta_2}\\\\
+%	\text{dom}(\theta_1) \cap \text{dom}(\theta_2) = \emptyset
+%}{
+%	\opType{\tvarCtx}{\fvalCtx}{\theta_1; \theta_2}{\Theta_1, \Theta_2}
+%}
+%\end{mathpar}
+%$\fbox{$\exprOKX{e}$}$
+%\begin{mathpar}
+%\inferrule[k-e-var]{ }{
+%	\exprOK{\tvarCtx}{\fvalCtx}{\evar{x}}
+%}
+%
+%\inferrule[k-e-lam]{
+%	\tKindX{\tau}{\kTypeBlur}\\
+%	\exprOK{\tvarCtx}{\fvalCtx}{e}
+%}{
+%	\exprOKX{\elam{\evar{x}}{\tau}{e}}
+%}
+%
+%\inferrule[k-e-op]{
+%	\fval{tycon}{-}{-}{\theta} \in \fvalCtx\\
+%	\tops{op}{\kappaidx}{i}{a}{-} \in \theta\\
+%	\tKindX{\tauidx}{\kappaidx}\\
+%	\exprOKX{e_1}\\
+%	\cdots\\
+%	\exprOKX{e_n}
+%}{
+%	\exprOKX{\eop{tycon}{op}{\tauidx}{\splat{e}{1}{n}}}
+%}
+%\end{mathpar}
+%$\fbox{$\kEq{\kappa}$}$
+%\begin{mathpar}
+%\inferrule[t-eq]{ }{
+%	\kEq{\kTypeBlur}
+%}
+%
+%\inferrule[i-eq]{ }{
+%	\kEq{\kint}
+%}
+%
+%\inferrule[l-eq]{ }{
+%	\kEq{\klabel}
+%}
+%
+%\inferrule[list-eq]{
+%	\kEq{\kappa}
+%}{
+%	\kEq{\klist{\kappa}}
+%}
+%\\
+%\inferrule[u-eq]{ }{
+%	\kEq{\kunit}
+%}
+%
+%\inferrule[p-eq]{
+%	\kEq{\kappa_1}\\
+%	\kEq{\kappa_2}
+%}{
+%	\kEq{\kpair{\kappa_1}{\kappa_2}}
+%}
+%
+%\inferrule[s-eq]{
+%	\kEq{\kappa_1}\\
+%	\kEq{\kappa_2}
+%}{
+%	\kEq{\ksum{\kappa_1}{\kappa_2}}
+%}
+%\end{mathpar}
+%\caption{\small Kinding for type and operator constructors and external terms, and equality kinds (see text).
+Type-level terms stored in the constructor context are not needed during kinding, indicated using a dash.}
+%\label{kindprog}
+%\vspace{-10pt}
+%\end{figure}
+
+%\section{Interactive Compilation and Invocation}\label{targets}
+%\subsection{OpenCL as an Active Library}
+%The code in this section uses \verb|clx|, an example library implementing the semantics of the OpenCL programming language and extending it with some additional useful types, which we will discuss shortly.
+Ace itself has no built-in support for OpenCL.
+%
+%To briefly review, OpenCL provides a data-parallel SPMD programming model where developers define functions, called {\em kernels}, for execution on \emph{compute devices} like GPUs or multi-core CPUs \cite{opencl11}.
+Each thread executes the same kernel but has access to a unique index, called its \emph{global ID}.
+Kernel code is written in a variant of C99 extended with some new primitive types and operators, which we will introduce  as needed in our examples below.
+%
+%\subsection{Generic Functions}\label{genfn}
+%%\begin{codelisting}[t]
+%%\lstinputlisting[commentstyle=\color{mauve}]{listing7.py}
+%%\caption{[\texttt{listing\ref{metaprogramming}.py}] Metaprogramming with Ace, showing how to construct generic functions from abstract syntax trees.}
+%%\label{metaprogramming}
+%%\end{codelisting}
+%
+%%\begin{codelisting}
+%%\lstinputlisting[linebackgroundcolor={\btLstHL{5-8}}]{listing3.py}
+%%\caption{[\texttt{listing\ref{map}.py}] A generic data-parallel higher-order map function targeting OpenCL.}
+%%\label{map}
+%%\end{codelisting}
+%%\begin{codelisting}[t]
+%%\lstinputlisting[commentstyle=\color{mauve}]{listing7.py}
+%%\caption{[\texttt{listing\ref{metaprogramming}.py}] Metaprogramming with Ace, showing how to construct generic functions from abstract syntax trees.}
+%%\label{metaprogramming}
+%%\end{codelisting}
+%Lines 3-4 introduce \verb|map|, an Ace function of three arguments that is governed by the \emph{active base} referred to by \verb|clx.base| and targeting the \emph{active target} referred to by \verb|clx.opencl|.
+The active target determines which language the function will compile to (here, the OpenCL kernel language) and mediates code generation.
+%
+%The body of this function, highlighted in grey for emphasis, does not have Python's semantics.
+Instead, it will be governed by the active base together with the active types used within it.
+No such types have been provided explicitly, however.
+Because our type system is extensible, the code inside could be meaningful for many different assignments of types to the arguments.
+We call functions awaiting types \emph{generic functions}.
+Once types have been assigned, they are called \emph{concrete functions}.
+%
+%Generic functions are represented at compile-time as instances of \verb|ace.GenericFn| and consist of an abstract syntax tree, an {active base} and an {active target}.
+The purpose of the \emph{decorator} on line 3 is to replace the Python function on lines 4-8 with an Ace generic function having the same syntax tree and the provided active base and active target.
+%Decorators in Python are simply syntactic sugar for applying the decorator function directly to the function  being decorated \cite{python}.
+In other words, line 3 could be replaced by inserting the following  statement on line 9:
+%\vspace{-0.28cm}
+%\begin{verbbox}
+%map = ace.fn(clx.base, clx.opencl)(map)
+%\end{verbbox}
+%\begin{figure}[h!]
+%\centering
+%\theverbbox
+%\end{figure}
+%\vspace{-0.28cm}
+%
+%The abstract syntax tree for \verb|map| is extracted using the Python standard  library packages  \verb|inspect| (to retrieve its source code) and \verb|ast| (to parse it into a syntax tree). 
+%
+%%\subsection{Metaprogramming in Ace}
+%%Generic functions can be generated directly from ASTs as well, providing Ace with support for  straightforward metaprogramming.
+Listing \ref{metaprogramming} shows how to generate two more generic functions, \verb|scale| and \verb|negate|.
+The latter is derived from the former by using a library for manipulating Python syntax trees, \verb|astx|.
+In particular, the \verb|specialize| function replaces uses of the second argument of \verb|scale| with the literal \verb|-1| (and changes the function's name), leaving a function of one argument.
+% 
+%\subsection{Concrete Functions and Explicit Compilation}
+%To compile a generic function to a particular \emph{concrete function}, a type must be provided for each argument, and typechecking and translation must then succeed.
+Listing \ref{compscript} shows how to explicitly provide type assignments to \verb|map| using the subscript operator (implemented using Python's operator overloading mechanism).
+We attempt to do so three times in Listing \ref{compscript}.
+The first, on line \ref{compscript}.7, fails due to a type error, which we handle so that the script can proceed.
+The error occurred  because the ordering of the argument types was incorrect.
+We provide a valid ordering on line \ref{compscript}.9 to generate the concrete function \verb|map_neg_f32|.
+We then provide a different type assignment to generate the concrete function \verb|map_neg_ci32|.
+%Concrete functions are instances of \verb|ace.TypedFn|, consisting of an abstract syntax tree annotated with types and translations along with a reference to the original generic function.
+%The typing and translation process is mediated by the logic in the active base, types and target that have been provided, as we will describe in more detail below.
+%
+%To produce an output file from an Ace ``compilation script'' like \verb|listing|\texttt{\ref{compscript}}\verb|.py|, the command \verb|acec| can be invoked from the shell, as shown in Listing \ref{mapc}.
+%\subsection{Types}
+%\begin{codelisting}
+%\lstinputlisting{listing4.py}
+%\caption{[\texttt{listing\ref{compscript}.py}] The generic \texttt{map} function compiled to map the \texttt{negate} function over two  types of input.}
+%\label{compscript}
+%\end{codelisting}
+%\begin{codelisting}
+%\begin{lstlisting}[style=Bash]
+%$ `acec listing3.py`
+%Hello, compile-time world!
+%[ace] TypeError in listing1.py (line 6, col 28): 
+%      'GenericFnType(negate)' does not support [].
+%[acec] listing3.cl successfully generated.
+%\end{lstlisting}
+%\caption{Compiling \texttt{listing\ref{compscript}.py} using the \texttt{acec} compiler.}
+%\label{mapc}
+%\end{codelisting}
+%\begin{codelisting}
+%\lstinputlisting[style=OpenCL]{listing5.cl}
+%\caption{[\texttt{listing\ref{compscript}.cl}] The OpenCL file generated by Listing \ref{mapc}.}
+%\label{mapout}
+%\end{codelisting}
+%%Lines \ref{compscript}.3-\ref{compscript}.5 construct the types assigned to the arguments of \verb|map| on lines \ref{compscript}.7-\ref{compscript}.10.
+In Ace, types are themselves values that can be manipulated at compile-time.
+This stands in contrast to other contemporary languages, where user-defined types (e.g.
+datatypes, classes, structs) are written declaratively at compile-time but cannot be constructed, inspected or passed around programmatically.
+More specifically, types are instances of a Python class that implements the \verb|ace.ActiveType| interface (see Sec.
+\ref{atypes}).
+%%As Python values, types can be assigned to variables when convenient (removing the need for  facilities like \verb|typedef| in C or \verb|type| in Haskell).
+Types, like all compile-time objects derived from Ace base classes, do not have visible state and operate in a referentially transparent manner (by constructor memoization, which we do not detail here).% These types are all implemented in the \verb|clx| library imported on line 1, none are built into Ace itself.
+%
+%The type named \verb|T1| on line \ref{compscript}.3 corresponds to the OpenCL type \verb|global float*|: a pointer to a 32-bit floating point number stored in the compute device's global memory (one of four address spaces defined by OpenCL \cite{opencl11}).
+It is constructed by applying \verb|clx.Ptr|, which is an Ace type constructor corresponding to pointer types, to a value representing the  address space, \verb|clx.global_|, and the type being pointed to.
+That type, \verb|clx.float|, is in turn the Ace type corresponding to \verb|float| in OpenCL (which, unlike C99, is always 32 bits).
+%The \verb|clx| library contains a full implementation of the OpenCL type system (including behaviors, like promotions, inherited from C99).
+%Ace is \emph{unopinionated} about issues like memory safety and the wisdom of such promotions.
+We will discuss how to implement, as libraries, abstractions that are higher-level than raw pointers in Sec.
+\ref{examples}, but Ace does not prevent users from choosing a low level of abstraction or ``interesting'' semantics if the need arises (e.g.
+for compatibility with existing libraries; see the discussion in Sec.
+\ref{discussion}).
+We also note that we are being more verbose than necessary for the sake of pedagogy.
+The \verb|clx| library includes more concise shorthand for OpenCL's types: \verb|T1| is equal to \verb|clx.gp(clx.f32)|.
+%Similarly, the decorators in Listings \ref{map} and \ref{metaprogramming} could have been written \verb|clx.cl_fn|.\todo{move this back there probably}
+%
+%The type \verb|T2| on line \ref{compscript}.4 is a pointer to a \emph{complex integer} in global memory.
+It does not correspond direrctly to a type in OpenCL, because OpenCL does not include primitive support for complex numbers.
+Instead, it uses an active type constructor \verb|clx.Cplx|, which includes the necessary logic for typechecking operations on complex numbers and translating them to OpenCL (Sec.
+\ref{atypes}).
+This constructor is parameterized by the numeric type that should be used for the real and imaginary parts, here \verb|clx.int|, which corresponds to 32-bit OpenCL integers.
+Arithmetic operations with other complex numbers, as well as with plain numeric types (treated as if their imaginary part was zero), are supported.
+When targeting OpenCL, Ace expressions assigned type \verb|clx.Cplx(clx.int)| are compiled to OpenCL expressions of type \verb|int2|, a  \emph{vector type} of two 32-bit integers (a type that itself is not inherited from C99).
+This can be observed in several places on lines \ref{mapout}.14-\ref{mapout}.21.
+This choice is merely an implementation detail that can be kept private to \verb|clx|, however.
+An Ace value of type \verb|clx.int2| (that is, an actual OpenCL vector) \emph{cannot} be used when a \verb|clx.Cplx(clx.int)| is expected (and attempting to do so will result in a static type error): \verb|clx.Cplx| truly extends the type system, it is not a type alias.
+%
+%The type \verb|TF| on line \ref{compscript}.5 is extracted from the generic function \verb|negate| constructed in Listing \ref{metaprogramming}.
+Generic functions, according to Sec.
+\ref{genfn}, have not yet had a type assigned to them, so it may seem perplexing that we are nevertheless assigning a type to \verb|negate|.
+Although a conventional arrow type cannot be assigned to \verb|negate|, we can give it a \emph{singleton type}: a type that simply means ``this expression is the \emph{particular} generic function \verb|negate|''.
+This type could also have been explicitly written as \verb|ace.GenericFnType(listing2.negate)|.
+During typechecking and translation of \verb|map_neg_f32| and \verb|map_neg_ci32|, the call to \verb|f| on line \ref{map}.6 uses the type of the provided argument to compile the generic function that inhabits the singleton type of \verb|f| (\verb|negate| in both of these cases) to a concrete function.
+This is why there are two versions of \verb|negate| in the output in Listing \ref{mapout}.
+In other words, types \emph{propagate} into generic functions -- we didn't need to compile \verb|negate| explicitly.
+This also explains the error printed on line \ref{mapc}.3-\ref{mapc}.4: when this type was inadvertently assigned to the first argument \verb|input|, the indexing operation on line \ref{map}.6 resulted in an error.
+A generic function can only be \emph{statically} indexed by a list of types to turn it into a concrete function, not \emph{dynamically} indexed with a value of type \verb|clx.size_t| (the return type of the OpenCL primitive function \verb|get_global_id|).
+%
+%In effect, this scheme enables higher-order functions even when targeting languages, like OpenCL, that have no support for higher-order functions (OpenCL, unlike C99, does not support function pointers).
+Interestingly, because they have a singleton type, they are higher-order but not first-class functions.
+That is, the type system would prevent you from creating a heterogeneous list of generic functions.
+Concrete functions, on the other hand, can be given both a singleton type and a true function type.
+For example, \verb|listing2.negate[[clx.int]]| could be given type \verb|ace.Arrow(clx.int, clx.int)|.
+The base determines how to convert the Ace arrow type to an arrow type in the target language (e.g.
+a function pointer for C99, or an integer that indexes into a jump table constructed from knowledge of available functions of the appropriate type in OpenCL).
+%
+%Type assignment to generic functions is similar in some ways to template specialization in C++.
+In effect, both a template header and type parameters at call sites are being generated automatically by Ace.
+This simplifies a sophisticated feature of C++ and enables its use with other targets like OpenCL.
+%Other uses for C++ templates (and the preprocessor) are subsumed by the metaprogramming features discussed above\todo{cite template metaprogramming paper}.
+%
+%%\subsection{Within-Function Type Resolution}
+%%\begin{codelisting}
+%%\lstinputlisting{listing6.py}
+%%\caption{\texttt{[listing6.py]} A function demonstrating whole-function type inference when multiple values with differing types are assigned to a single identifier, \texttt{y}.}
+%%\label{inference}
+%%\end{codelisting}
+%%On line 5 in the generic \verb|map| function in Listing \ref{map}, the variable \verb|gid| is initialized with the result of calling the OpenCL primitive \verb|get_global_id|.
+The type for \verb|gid| is never given explicitly.
+This is a simple case of Ace's {\em within-function type resolution} strategy (we hesitate to call it \emph{type inference} because it does not, strictly speaking, take a constraint-solving approach).
+In this case, the type of \verb|gid| will resolve to \verb|size_t| because that is the return type of \verb|get_global_id| (as defined in the OpenCL specification, which the \verb|ace.OpenCL| module follows).
+The result can be observed on Lines 11 and 24 in Listing \ref{mapout}.
+%%
+%%Inference is not restricted within single assignments, as in the \verb|map| example, however.
+Multiple assignments to the same identifier with values of differing types, or multiple return statements, can be combined if the types in each case are compatible with one another (e.g.
+by a subtyping relation or an implicit coercion).
+In Listing \ref{inference}, the \verb|threshold_scale| function assigns different values to \verb|y| in each branch of the conditional.
+In the first branch, the value \verb|0| is an \verb|int| literal.
+However, in the second branch of the loop, the type depends on the types of both arguments, \verb|x| and \verb|scale|.
+We show two choices for these types on Lines 11 and 12.
+Type inference correctly combines these two types according to OpenCL's C99-derived rules governing numeric types (defined by the user in the \verb|OpenCL| module, as we will describe in Section \ref{att}).
+We can verify this programmatically on Lines 12 and 13.
+Note that this example would also work correctly if the assignments to \verb|y| were replaced with \verb|return| statements (in other words, the return value of a function is treated as an assignable for the purpose of type inference).
+%
+%\subsection{Implicit Compilation and Interactive Execution}\label{compenv}\label{backend}\label{implicit}
+%%Professional end-user programmers (e.g.
+scientists and engineers \cite{professional-end-users}) today generally use dynamically-typed high-level languages like MATLAB, Python, R or Perl for tasks that are not performance-sensitive, such as small-scale data analysis and plotting \cite{nguyen2010survey}.
+For portions of their analyses where the performance overhead of dynamic type checking and automatic memory management is too high, they will typically call into code written in a statically-typed, low-level language, most commonly C or Fortran, that uses low-level parallel abstractions like pthreads and MPI \cite{4222616,basili2008understanding}.
+Unfortunately, these low-level languages and abstractions are notoriously difficult to use and automatic verification is intractable in general.\todo{integrate this}
+%
+%
+%A common workflow for \emph{professional end-user programmers} (e.g.
+scientists and engineers) is to use a simple scripting language for orchestration, small-scale data analysis and visualization and call into a low-level language for performance-critical sections.
+Python is both designed for this style of use and widely adopted for such tasks \cite{sanner1999python,nguyen2010survey}.
+Developers can call into native functions using Python's foreign function interface (FFI), for example.
+A more recent trend is to generate and compile code without leaving Python, using a Python wrapper around a compiler.
+For example, \verb|weave| works with C and C++, and \verb|pycuda| and \verb|pyopencl| work with CUDA and OpenCL, respectively \cite{klockner2011pycuda}.
+%\begin{codelisting}
+%\lstinputlisting{listing8.py}
+%\caption{[\texttt{listing\ref{py}.py}] A full OpenCL program using the \texttt{clx} Python bindings, including data transfer to and from a device and direct invocation of a generic function, \texttt{map}.}
+%\label{py}
+%\end{codelisting}
+%The OpenCL language was designed for this workflow, exposing a retargetable compiler and data management  routines as an API, called the \emph{host API} \cite{opencl11}.
+The \verb|pyopencl| library exposes this API to Python and simplifies interoperation with \verb|numpy|, a popular package for manipulating contiguously-allocated numeric arrays in Python \cite{klockner2011pycuda}.
+%
+%Ace supports a refinement to this workflow, as an alternative to the \verb|acec| compiler described above, for targets that have wrappers like this available, including \verb|clx.opencl|.
+Listing \ref{py} shows an example of this workflow where the user chooses a compute device (line \ref{py}.3),  constructs a \verb|numpy| array (line \ref{py}.5), transfers it to the device (line \ref{py}.6), allocates an empty equal-sized buffer for the result of the computation (line \ref{py}.7), launches the {generic} kernel \verb|map| from Listing \ref{map} with these device arrays as well as the function \verb|negate| from Listing \ref{metaprogramming} (line \ref{py}.9) choosing a number of threads equal to the number of elements in the input array (line \ref{py}.10), and transfers the result back into main memory to check that the device computed the expected result (line \ref{py}.12).
+%
+%For developers experienced with the usual OpenCL or CUDA workflow, the fact that this can be accomplished in a total of 6 statements may be surprising.
+This simplicity is made possible by Ace's implicit tracking of types throughout the code.
+First, \verb|numpy| keeps track of the type, shape and order of its arrays.
+The type of \verb|input|, for example, is \verb|numpy.float64| by default, its shape is \verb|(1024,1024)| and its order is row-major by default.
+The \verb|pyopencl| library, which the \verb|clx| mechanism is built upon, uses this metadata to automatically call the underlying OpenCL host API function for transferring byte arrays to the device without requiring the user to calculate the size.
+The \verb|clx| wrapper further retains this metadata in \verb|d_input| and \verb|d_output|, the Python wrappers around the allocated device arrays.
+The Ace active type of these wrappers is an instance of \verb|clx.NPArray| parameterized by this metadata.
+This type knows how to typecheck and translate operations, like indexing with a multi-dimensional thread index, automatically.
+%
+%By extracting the types from the arguments, we can call the generic function \verb|map| without first requiring an explicit type assignment, like we needed when using \verb|acec| above.
+In other words, dynamic types and other metadata can propagate from Python data structures into an Ace generic function as static type information, in the same manner as it propagated \emph{between} generic functions in the previous section.
+In both cases, typechecking and translation of \verb|map| happens the first time a particular type assignment is encountered and cached for subsequent use.
+When called from Python, the generated OpenCL source code is compiled for the device we selected using the OpenCL retargetable compilation infrastructure, and cached for subsequent calls.
+%
+%The same program written using the OpenCL C API directly is an order of magnitude longer and significantly more difficult to comprehend.
+OpenCL does not support higher-order functions nor is there any way to write \verb|map| in a type-generic or shape-generic manner.
+If we instead use the \verb|pyopencl| library and apply the techniques described in \cite{klockner2011pycuda}, the program is still twice as large and less readable than this code.
+Both the \verb|map| and \verb|negate| functions must be explicitly specialized with the appropriate types using string manipulation techniques, and custom shapes and orders can be awkward to handle.
+Higher order functions are still not available, and must also be simulated by string manipulation.
+That approach also does not permit the use any of the language extensions that Ace enables (beyond the useful type for \verb|numpy| arrays just described; see Sec.
+\ref{examples} for more interesting possibilities).
+%
+%%Not shown are several additional conveniences, such as delegated kernel sizing and \verb|In| and \verb|Out| constructs that can reduce the size and improve the clarity of this code further; due to a lack of space, the reader is referred to the language documentation.
+
+\section{Related Work}\label{related}
+
+%\parindent0pt
+%
+\begin{figure*}
+\small
+\vspace{-10pt}
+\onecolumn
+\begin{longtable}{l l@{}l c@{}c c@{}c c@{}c c@{}c c@{}c}
+%\toprule%
+
+&&  &&  && {\bfseries Extensible} && {\bfseries Extensible} && {\bfseries Unambiguous} && {\bfseries Alternative}\\
+
+{\bfseries Approach} && {\bfseries Examples} && {\bfseries Library} && {\bfseries Syntax} && {\bfseries Type System} && {\bfseries Composition} && {\bfseries Targets}  \\
+
+\cmidrule(l){1-1} \cmidrule(l){2-3} \cmidrule(l){4-5} \cmidrule(l){6-7} \cmidrule(l){8-9} \cmidrule(l){10-11} \cmidrule(l){12-13}
+
+\endhead
+
+Active Types && Ace && \CIRCLE && \Circle && \CIRCLE && \CIRCLE && \CIRCLE \\
+
+\myrowcolour%
+Desugaring && SugarJ \cite{erdweg2011sugarj} && \Circle && \CIRCLE && \Circle && \Circle && \Circle \\
+Type-Specific Literals && Wyvern \cite{Omar:2013:TWP:2489812.2489815} && \Circle && \CIRCLE && \Circle && \CIRCLE && \Circle \\
+\myrowcolour%
+Rule Injection && Racket \cite{TypedScheme2008}, Xroma \cite{activelibraries}, mbeddr \cite{mbeddr} &&  \cite{TypedScheme2008} && \cite{mbeddr} && \CIRCLE && \Circle && \Circle \\
+Attribute Grammars && Silver \cite{VanWyk:2010:SEA}, Xoc \cite{conf/asplos/CoxBCKK08} && \Circle && \CIRCLE && (unsafe)&& \cite{conf/pldi/SchwerdfegerW09} && \Circle\\
+\myrowcolour
+Static macros /  && Scala \cite{ScalaMacros2013}, MorphJ \cite{MorphJ2011}, && \Circle && \Circle && \Circle && \CIRCLE && \Circle \\
+\myrowcolour
+Metaprogramming &&  MetaML \cite{Sheard:1999:UMS}, Template Haskell && && && && && \\
+
+Cross-Compilation && LMS/Delite \cite{Rompf:2012:LMS, Delite2011} && \CIRCLE && \Circle && \Circle && \Circle && \CIRCLE \\
+
+\myrowcolour%
+%EDSL Frameworks && ? && \CIRCLE && \CIRCLE && \CIRCLE && \Circle && \Circle \\
+\end{longtable}
+\caption{Comparison to related approaches to language extensibility.}\label{relatedtable}
+\twocolumn
+\vspace{-10pt}
+\end{figure*}
+%\subsection{Active Libraries}
+
+%We have implemented the full OpenCL programming language as  primary example showed how to support low-level GPU programming from a high-level language.
+Other languages have supported similar workflows, e.g.
+Rust \cite{rustgpu}.
+These have not been fundamentally extensible and suffer from the ``bootstrapping'' problem described in Sec.
+\ref{intro}.
+
+Libraries containing compile-time logic have previously been called {\it active libraries} \cite{activelibraries}.
+A number of  projects, such as Blitz++, have taken advantage of C++ template metaprogramming to implement domain-specific optimizations \cite{veldhuizen2000blitz++}.
+Others have chosen custom metalanguages (e.g.
+Xroma \cite{activelibraries}, mbeddr \cite{mbeddr}).
+In Ace, we replace these brittle mini-languages with a general-purpose language, Python, and significantly expand the notion of active libraries by consideration of types, base semantics and target languages as values (in this case, objects).
+We borrow the term ``active'' due to this relationship.
+
+%Generic functions are a novel strategy for {\it function polymorphism} -- defining functions that operate over more than a single type.
+In Ace, generic functions are implicitly polymorphic and can be called with arguments of {\it any type that supports the operations used by the function}.
+This is related to structural polymorphism \cite{malayeri2009structural}.
+Structural types make explicit the structure required of an argument, unlike generic functions, which are only given singleton types because the structure may depend on the semantics of an active type.
+%Structural typing can be compared to the more \emph{ad hoc} approach taken by dynamically-typed languages like Python itself, sometimes called ``duck typing''.
+It is also comparable to the C++ template system, as discussed previously.
+
+{\it Operator overloading} \cite{vanWijngaarden:Mailloux:Peck:Koster:Sintzoff:Lindsey:Meertens:Fisker:acta:1975} and {\it metaobject dispatch} \cite{Kiczales91} are run-time protocols that translate operator invocations into function calls.
+The function is typically selected according to the dynamic type or value of one or more operands.
+These protocols share the notion of {\it inversion of control} with our strategy.
+However, our strategy is a {\it compile-time} protocol where the typechecking and translation semantics are implemented for different operators.
+Note that for syntactic convenience, we used Python's operator overloading and metaobject protocol substantially at the type level.%Our dispatch protocol was designed to be conceptually similar to Python's operator overloading protocol \cite{python}.
+
+There are several other mechanisms that have been described as enabling forms of internal ``extension'' in the literature.
+These differ from our mechanism in one of the following ways, summarized in Fig.
+\ref{relatedtable}:
+\begin{itemize}
+\item Our mechanism is itself implemented as a library, rather than as a language dialect.
+\item Our mechanism reuses an existing language's syntax, rather than offering syntax extension support.
+As we saw, this can still be reasonably natural, and allows the language to benefit from a variety of existing tools.
+We avoid many facets of the expression problem \cite{wadler1998expression} in this way.
+\item Our mechanism permits true type system extensions.
+New types are not merely aliases, nor must their rules be admissible in some base type system.
+Term rewriting and macro systems, as well as systems that involve cross-compilation do not handle type system extensions at all.
+LMS, for example, uses Scala's type system, focusing instead on code generation and optimization \cite{Rompf:2012:LMS}.
+This is orthogonal to the goals we are seeking to achieve (e.g.
+the regular expression types we have described would be somewhat difficult to implement as libraries in Scala).
+\item Techniques that allow global extensions to the syntax or treat the semantics as a ``bag of rules'' (e.g.
+SugarJ or rule injection systems like Xroma, Typed Racket (and other typed LISPs) and mbeddr) allow extensions so much control that there can be ambiguities.
+In Ace, there \emph{can never be more than one extension assigned as the delegate for a term}, so there are no ambiguities.
+\item Our mechanism allows users to control the target language of compilation from within libraries.
+\end{itemize}
+
+When the mechanisms available in an existing language prove insufficient, researchers and domain experts often design a new language.
+Previous work has considered adding statically-typed features to languages like Python in this way.
+For example, Terra embeds a low-level statically-typed language within Lua \cite{terra}.
+A number of tools have been developed to assist with this task, including compiler generators, language workbenches and domain-specific language frameworks (cf.
+\cite{erdweg2013state}).
+Extensible compilers like Xoc \cite{conf/asplos/CoxBCKK08} can be considered a form of language framework as well.
+It is difficult or impossible for these language-external approaches to achieve interoperability, because languages are not aware of each other's semantics and merging languages is not guaranteed to be unambiguous and sound.
+A few approaches (e.g.
+) have made steps toward addressing the issue of ambiguity (albeit with some cost in expressiveness), but soundness is not guaranteed.
+In Ace (and in the theory, assuming a suitable packaging system), using different type constructors at the same time cannot cause type safety issues because representational consistency is checked.
+Representational consistency permits compositional reasoning.
+
+Bidirectional type systems have been used for adding refinement checking to languages like ML and Twelf \cite{Lovas08abidirectional}.
+Refinement types can add stronger static logic, but do not have control over translation and have not been shown practical in a language as widely adopted as Python.
+We believe this is a fruitful avenue for future work.
+The Wyvern programming language uses bidirectional typechecking to control aspects of parsing in a manner similar to how we treat introductory forms, but Wyvern is also a language dialect and does not have an extensible type system \cite{Omar:2013:TWP:2489812.2489815}.
+
+\section{Discussion}\label{discussion}
+This work aimed to show that one can add static typechecking to a language like Python as a library, without undue syntactic overhead.
+Moreover, the type system is not fixed, but flexibly extensible.
+We achieve this by using a bidrectional type system and an elaboration semantics targeting a user-defined target language, along with per-function base semantic annotations.
+We show that the core of this mechanism leads to a surprisingly clean theoretical formalism that combines work on type-level computation, typed compilation and bidirectional type systems to enable type system extensions over a fixed, but flexible, syntax.
+
+% Different dialects can be difficult to combine manually, as we discussed in the introduction.
+We have implemented the full OpenCL programming language (a variant of C99) as a library (though the details are left for future work), showing that such languages can instead be written as libraries, rather than as dialects.
+% 
+%\subsection{Compositional Reasoning}\label{safety}
+%Every statement and expression in an Ace function is governed by exactly one active type, determined by the dispatch protocol, or by the single active base associated with the Ace function.
 The representational consistency check ensures that compilation is successful without requiring that types that abstract over other types (e.g.
@@
 
-\section{Theoretical Foundations}\label{theory}
-\begin{figure}
-\small
-\begin{flalign}
-\small
- & \family{Tuple}{\klist{\kTypeBlur}}{\\
- & \quad \tops{new}{\kunit}{\_}{a}{\tlam{\_}{\kunit}{\tlam{a}{\klist{\kDen}}{\\
- & \quad\quad
- 	\tfold{\tvar{a}}{\tden{{()}}{\ttype{Tuple}{\tnil{\kTypeBlur}}}}{d}{b}{r}{\\
-			& \quad\quad\quad \tfamcase{\ttypeof{\tvar{r}}}{Ntuple}{i}{\\
-			& \quad\quad\quad\quad \tfold{\tvar{b}}{\tden{\itransof{\tvar{d}}}{\ttype{Ntuple}{\tcons{\ttypeof{\tvar{d}}}{\tvar{i}}}}\\& \quad\quad\quad\quad}{\_}{\_}{\_}{\tden{\\& \quad\quad\quad\quad\quad {(\itransof{\tvar{d}},\itransof{\tvar{r}})}}{\ttype{Ntuple}{\tcons{{\ttypeof{\tvar{d}}}}{\tvar{i}}}}}
-			}{\terr}
-	}
- }}};\\
- & \quad \tops{prj}{\kint}{i}{a}{\lambda \tvar{i}:\kint.\lambda \tvar{a}:\klist{\kDen}. 
- 	\tapp{\tvar{arity1}}{\tapp{\tvar{a}}{\tlam{d}{\kDen}{\\
-	& \quad\quad\tfamcase{\ttypeof{\tvar{d}}}{Ntuple}{nl}{\\
- & \quad\quad \tfold{\tvar{nl}}{\terr}{t1}{j}{\_}{\\
- & \quad\quad \tfold{\tvar{j}}{\tifeq{\tvar{i}}{1}{\dint}{\tden{\itransof{d}}{\tvar{t1}}}{\terr}\\&\quad\quad\quad}{\_}{\_}{\_}{\\
- & \quad\quad\quad (\tlam{p}{\kpair{\kDen}{\dint}}{\tifeq{\tsnd{\tvar{p}}}{\tvar{i}}{\dint}{\tfst{\tvar{p}}}{\terr}})\\
- & \quad\quad\quad (\tapp{\tvar{foldl}}{\tapp{\tvar{nl}}{\tapp{(\tden{\tvar{x}}{\tvar{nt}},0)}{
- 	\tlam{r}{\kpair{\kDen}{\dint}}{\tlam{t}{\kTypeBlur}{\tlam{ts}{\klist{\kTypeBlur}}{\\
-	& \quad\quad\quad\quad \tifeq{\tvar{i}}{\tsnd{\tvar{r}}}{\dint}{\tvar{r}}{\tdencase{\tfst{\tvar{r}}}{rx}{\_}{\\
-	& \quad\quad\quad\quad\quad \tifeq{\tvar{i}}{\tsnd{\tvar{r}}+1}{\dint}{\\
-	& \quad\quad\quad\quad\quad\quad \tifeq{\tvar{ts}}{\tnil{\kTypeBlur}}{\klist{\kTypeBlur}}{(\tden{\tvar{rx}}{\tvar{t}},\tvar{i})\\
-	& \quad\quad\quad\quad\quad\quad}{(\tden{\titerm{\ifst{\iup{\tvar{rx}}}}}{\tvar{t}},\tvar{i})}\\	
-	& \quad\quad\quad\quad\quad}{(\tden{\titerm{\isnd{\iup{\tvar{rx}}}}}{\tvar{t}},\tsnd{\tvar{r}}+1)}}{\terr}}}}}
- }}})
- }}{\terr}}}}}
- }\\ & 
-%& \quad \tops{pr}a}{b}{c}{d}
- }{d}{i}{(\tlam{i}{\klist{\kTypeBlur}}{
- 	\\& \quad \quad \tfold{\tvar{i}}{\titype{\dunit}}{s}{j}{r}{
- 		\tfold{\tvar{j}}{\titype{\trepof{\tvar{s}}}}{\_}{\_}{\_}{\\
-		&\quad\quad\quad
- 		\titype{\dpair{\trepof{\tvar{s}}}{\dup{\tvar{r}}}}
- 		}
- 	}
- })}
-\end{flalign}
-\caption{ABC}
-\label{tuple}
-\end{figure}
-\begin{figure}[t]
-\small
-$\fbox{\inferrule{}{\progOKX{\progsort}}}$
-~~~$\tvarCtx ::= \emptyctx \pipe \tvarCtxX{t}{\kappa}$
-~~~$\fvalCtx ::= \emptyctx \pipe \fvalCtxX{\fvalDf}$
-%$\fCtx ::= \Sigma_0	 \pipe \fvalCtxX$
-\begin{mathpar}
-\inferrule[k-tycon]{
-	\fvar{Tycon} \notin \text{dom}(\fvalCtx)\\
-	\kEq{\kappaidx}\\
-	\tKind{\tvarCtx}{\fvalCtx}{\taurep}{\karrow{\kappaidx}{\kIType}}\\
-	\opType{\tvarCtx}{\fvalCtxX{\fvalDf}}{\theta}{\Theta}\\
-	\progOK{\tvarCtx}{\fvalCtxX{\fvalDf}}{\rho}
-}{
-	\progOKX{\pfam{\familyDf}{\rho}}
-}
-
-%\inferrule[def-kinding]{
-%	\tKindX{\tau}{\kappa}\\
-%	\progOK{\tvarCtxX{t}{\kappa}}{\fvalCtx}{\rho}
-%}{
-%	\progOKX{\pdef{t}{\kappa}{\tau}{\rho}}
-%}
-%
-\inferrule[k-e-prog]{
-	\exprOK{\tvarCtx}{\fvalCtx}{e}
-}{
-	\progOKX{e}
-}
-\end{mathpar}
-$\fbox{$\opType{\tvarCtx}{\fvalCtx}{\theta}{\Theta}$}$
-\begin{mathpar}
-\inferrule[k-opcon]{
-	\tKind{\tvarCtx}{\fvalCtx}{\taudef}{\karrow{\kappaidx}{\karrow{\klist{\kDen}}{(\ksum{\kDen}{\kunit})}}}
-}{
-	\opType{\tvarCtx}{\fvalCtx}{{\tops{op}{\kappaidx}{i}{a}{\taudef}}}{
-	\kOpS{op}{\kappaidx}}
-}
-
-\inferrule[k-opcons]{
-	\opType{\tvarCtx}{\fvalCtx}{\theta_1}{\Theta_1}\\
-	\opType{\tvarCtx}{\fvalCtx}{\theta_2}{\Theta_2}\\\\
-	\text{dom}(\theta_1) \cap \text{dom}(\theta_2) = \emptyset
-}{
-	\opType{\tvarCtx}{\fvalCtx}{\theta_1; \theta_2}{\Theta_1, \Theta_2}
-}
-\end{mathpar}
-$\fbox{$\exprOKX{e}$}$
-\begin{mathpar}
-\inferrule[k-e-var]{ }{
-	\exprOK{\tvarCtx}{\fvalCtx}{\evar{x}}
-}
-
-\inferrule[k-e-lam]{
-	\tKindX{\tau}{\kTypeBlur}\\
-	\exprOK{\tvarCtx}{\fvalCtx}{e}
-}{
-	\exprOKX{\elam{\evar{x}}{\tau}{e}}
-}
-
-\inferrule[k-e-op]{
-	\fval{Tycon}{-}{-}{\theta} \in \fvalCtx\\
-	\tops{op}{\kappaidx}{i}{a}{-} \in \theta\\
-	\tKindX{\tauidx}{\kappaidx}\\
-	\exprOKX{e_1}\\
-	\cdots\\
-	\exprOKX{e_n}
-}{
-	\exprOKX{\eop{Tycon}{op}{\tauidx}{\splat{e}{1}{n}}}
-}
-\end{mathpar}
-$\fbox{$\kEq{\kappa}$}$
-\begin{mathpar}
-\inferrule[t-eq]{ }{
-	\kEq{\kTypeBlur}
-}
-
-\inferrule[i-eq]{ }{
-	\kEq{\kint}
-}
-
-\inferrule[l-eq]{ }{
-	\kEq{\klabel}
-}
-
-\inferrule[list-eq]{
-	\kEq{\kappa}
-}{
-	\kEq{\klist{\kappa}}
-}
-\\
-\inferrule[u-eq]{ }{
-	\kEq{\kunit}
-}
-
-\inferrule[p-eq]{
-	\kEq{\kappa_1}\\
-	\kEq{\kappa_2}
-}{
-	\kEq{\kpair{\kappa_1}{\kappa_2}}
-}
-
-\inferrule[s-eq]{
-	\kEq{\kappa_1}\\
-	\kEq{\kappa_2}
-}{
-	\kEq{\ksum{\kappa_1}{\kappa_2}}
-}
-\end{mathpar}
-\caption{\small Kinding for type and operator constructors and external terms, and equality kinds (see text).
-Type-level terms stored in the constructor context are not needed during kinding, indicated using a dash.}
-\label{kindprog}
-\vspace{-10pt}
-\end{figure}
+%\begin{enumerate}
+%\item We address the ``chicken-and-egg'' problem by designing Ace, an extensible language  embedded within Python \cite{Politz:2013:PFM:2509136.2509536,python} entirely as a library.
+The top-level of an Ace file is a \emph{compilation script} written in Python and Python serves as Ace's \emph{type-level language}, as we will discuss.
+Ace and Python thus share a common syntax and package system, so users of Ace can directly leverage established tools, infrastructure and coding standards.
+Functions marked by a decorator as under the control of Ace are, unlike Python functions, \emph{statically} typechecked.
+This makes it possible to rule out many potential issues ahead of execution, and thus  avoid  costly dynamic checks.
+%The semantics of  define run-time behavior.
+%\item  % Instead, Ace has a statically-typed semantics that can be extended by users from within libraries. 
+%There is \emph{no particular type system built into Ace}.
+Instead, a type is any  object implementing the \verb|ace.Type| interface constructed by the compilation script.
+We call these \emph{active types}.
+When statically assigning a type to a compound form, e.g.
+\verb|e.x|, the Ace {compiler} delegates to the type of a subexpression, e.g.
+to the type of \verb|e|, to determine how to assign a type to the expression as a whole.
+As we will show, this {type-directed protocol}, \emph{active typing}, permits the expression of a rich variety of type systems as libraries.
+Unlike syntax-directed protocols, there cannot be ambiguities when separately defined type sytems are combined (types control non-overlapping sets of expressions by construction).
+Base cases, e.g.
+variables and certain statement forms, are handled on a {per-function basis} by a \emph{base semantics}, also a compile-time object implementing an interface, \verb|ace.Base|.
+%\end{enumerate}
+%% As a result, clients are able to import any combination of extensions with the confidence that link-time ambiguities cannot occur .
+%
+%The dynamic behavior of an Ace function is determined by translation to a \emph{target language}.
+We begin by  targeting Python, then discuss targeting OpenCL and CUDA, lower-level languages used to program many-core processors (e.g.
+GPUs).
+Active types and bases control translation by the same type-directed protocol that governs type assignment.
+We call this phase \emph{active translation}.
+A \emph{target} mediates this process and is also a user-defined compile-time object implementing an interface, \verb|ace.Target|.
+%%One base or type can support multiple targets. 
+%When targeting a typed language, care must be taken to ensure that well-typed Ace expressions have well-typed translations, so we integrate a technique for compositionally reasoning about compiler correctness (developed for the TIL compiler for Standard ML \cite{tarditi+:til-OLD}) into the language.
+%
+%%TODO
+%%
+%%Static type systems are powerful tools for programming language design and implementation.
+By tracking the type of a value statically, a typechecker can verify the absence of many kinds of errors over all inputs.
+This simplifies and increases the performance of the run-time system, as errors need not be detected dynamically using tag checks and other kinds of assertions.
+%
+%It is legitimate to ask, however, why dynamically-typed languages are so widely-used in multiple domains.
+Although slow and difficult to reason about, these languages generally excel at satisfying the criteria of \textbf{ease-of-use}.
+More specifically, Cordy identified the principle of \emph{conciseness} as elimination of
+%redundancy and the availability of reasonable defaults \cite{cordy1992hints}.
+Statically-typed languages, particularly those that professional end-users are exposed to, are often verbose, requiring explicit and often redundant type annotations on each function and variable declaration, separate header files, explicit template headers and  instantiation and other sorts of annotations.
+Dynamically-typed languages, on the other hand, avoid most of this overhead by relying on support from the run-time system.
+Ace was first conceived to explore the question: \emph{does conciseness require run-time mechanisms, or can one develop a statically-typed language with the same usability profile?}
+%
+%
+%Rather than designing a new syntax, or modifying the syntax of an existing language, we chose to utilize, \emph{without modification}, the syntax of an existing language, Python.
+This choice was not arbitrary, but rather a key means by which Ace satisfies extrinsic design criteria related to familiarity and tool support.
+Researchers often dismiss the importance of syntax.
+By repurposing a well-developed syntax, they no longer need to worry about the ``trivial'' task of implementing it.
+%
+%Most programming languages are {\em monolithic} -- a collection of primitives are given first-class treatment by the language implementation, and users can only creatively combine them to implement  abstractions of their design.
+Although highly-expressive general-purpose mechanisms have been developed (such as object systems or functional datatypes), these may not suffice when researchers or domain experts wish to evolve aspects of the type system, exert control over the representation of data, introduce specialized run-time mechanisms, if defining an abstraction in terms of existing mechanisms is unnatural or verbose, or if custom error messages are useful.
+In these situations, it would be desirable to have the ability to modularly extend existing systems with new compile-time logic and be assured that such extensions will never interfere with one another when used in the same program.
+%
+%Python does not truly prevent extensions from interfering with one another because it lacks, e.g., data hiding mechanisms.
+One type could change the implementation of another by replacing its methods, for example.
+But these kinds of conflicts do not occur ``innocently'', as conflicts between two libraries in SugarJ that use similar syntax might.
+% 
+%
+%
+%
+Ace has several limitations still.
+Debuggers and other tools that rely not just on Python's syntax but also its semantics cannot be used directly, so if code generation introduces significant complexity that leads to bugs, this can be an issue.
+We believe that active types can be used to control debugging, and plan to explore this in the future.
+We have also not yet evaluated the feasibility of implementing more advanced type systems (e.g.
+linear, dependent or flow-dependent type systems) using Ace.
+In particular, it is difficult to orthogonally implement type systems that use different contexts for typing, because the base controls the context (though hacks are possible).
+%
+Not all extensions will be useful.
+Indeed, some language designers worry that offering too much flexibility to users leads to abuse (this is, for example, widely credited as the reason why Java doesn't support operator overloading).
+We do not argue with this point.
+Instead, we argue that the potential for abuse must be balanced with the possibilities made available by a vibrant ecosystem of competing statically-typed abstractions that can be developed and deployed as libraries, and thus more easily evaluated in the wild.
+With an appropriate community process, this could lead to a convergence toward stable collections of curated, high-quality collections more quickly than is possible today.
+%We also anticipate that coding guidelines and tools mandating the use of abstractions that are known to have certain desirable properties will replace designer-mandated enforcement.
+% 
+%This paper is presented as a language design paper.
+The theoretical foundations of this work lie in type-level computation.
+We have developed a simplified, type-theoretic formalism of active typechecking and translation, where we prove the safety properties we have outlined here, as well as several additional ones that are only possible to achieve using a typed metalanguage (currently in submission at ESOP 2014).
+The work described here  extends that mechanism in several directions: Ace supports a rich syntax, makes syntax trees available to extensions, supports a pluggable backend and per-function base semantics.
+It is implemented in an existing language and supports a wider range of practical extensions.
+
+The mechanisms described here can be implemented within any language that offers some form of quotation of function ASTs, capturing of function closures, and a reasonably flexible collection of syntactic forms and basic reflection facilities.% It is particularly useful when the language supports interoperability layers with a number of target languages.
+
+%\acks
+%The author is grateful to Jonathan Aldrich, Robert Bocchino and anonymous referees for their useful suggestions.
+This work was funded by the DOE Computational Science Graduate Fellowship under grant number DE-FG02-97ER25308.
+%Acknowledgments, if needed.
+
+% We recommend abbrvnat bibliography style.
+
+\bibliographystyle{abbrv}
+
+% The bibliography should be embedded for final submission.
+
+\bibliography{../research}
+%\softraggedright
+%P. Q. Smith, and X. Y. Jones. ...reference text...
+
+\appendix
+\section{Source Code Emission}
+Portions of the \verb|Target| class related to source code emission are shown.
+If compilation occurred externally, the \verb|acec| library asks the targets used by the top-level typed functions to generate one or more files by deciding on a name and file extension on the basis of the name of the compilation script, and emitting code and \emph{snippets}, which are simply strings corresponding to functions, type declarations, imports and other top-level entities in the target language, each inserted at a \emph{location}.
+Each file being generated is a location, and locations might also be nested, depending on the language (e.g.
+the import block may be a separate location when generating Java, or the body of the \verb|<head>| tag when generating HTML+CSS+Javascript).
+Snippets are only added once to a particular location (so imports are not duplicated, for example).
+%We leave layering extensions to Python to created typed targets as future work.
+
+\begin{codelisting}
+\begin{lstlisting}
+class PyTarget(ace.Target):
+  class PyAST(object): """Base class for ASTs"""
+  target_type = ast
+  class Attribute(PyAST, ast.Attribute):
+    def emit(self, cg):
+      cg.append(self.value.emit(cg), '.', self.attr)
+  # ... 
+  class AnonModule(object): 
+    def __init__(self, name): 
+      self.name = name
+      self.guid = Guid()
+
+    def emit(self, cg):
+      import_loc = cg['imports']
+      _cache = AnonModule._idx_cache[import_loc]
+      if self not in _cache:
+        anon_name = '__ace_' + str(len(_cache)+1)
+        imp_stmt = ('import ' + self.name + ' as ' + 
+          anon_name + cg.newline)
+        _cache[self] = (anon_name, imp_stmt)
+      else: 
+        anon_name, imp_stmt = _cache[self]
+      imports.add_snippet(imp_stmt)
+      cg.append(anon_name)
+
+  def emit(self, script_name, cg):
+    file_name = '_' + script_name + '.py'
+    if file_name not in generator:
+      if 'imports' not in generator[file_name]:
+        generator[file_name].push_loc('imports')
+      if 'main' not in format_set[file_name]:
+        generator[file_name].push_loc('main')
+    translation.emit((file_name, 'main'))
+\end{lstlisting}
+\caption{Portions of the target showing how non-local code emission works.}
+\label{cg}
+\end{codelisting}
+
+\section{Kinding}
 \begin{figure}[t]
@@
 \inferrule[k-ty-i]{
-	\fval{Tycon}{\kappaidx}{-}{-} \in \fvalCtx\\\\
+	\fval{tycon}{\kappaidx}{-}{-} \in \fvalCtx\\\\
 	\tKind{\tvarCtx}{\fvalCtx}{\tauidx}{\kappaidx}
 }{
-	\tKindX{\ttype{Tycon}{\tauidx}}{\kTypeBlur}
+	\tKindX{\ttype{tycon}{\tauidx}}{\kTypeBlur}
 }
@@
 	\tKindX{\tau}{\kTypeBlur}\\
-	\fval{Tycon}{\kappaidx}{-}{-} \in \fvalCtx\\\\
+	\fval{tycon}{\kappaidx}{-}{-} \in \fvalCtx\\\\
 	\tKind{\tvarCtxX{x}{\kappaidx}}{\fvalCtx}{\tau_1}{\kappa}\\
@@
 }{
-	\tKind{\tvarCtx}{\fvalCtx}{\tfamcase{\tau}{Tycon}{x}{\tau_1}{\tau_2}}{\kappa}
+	\tKind{\tvarCtx}{\fvalCtx}{\tfamcase{\tau}{tycon}{x}{\tau_1}{\tau_2}}{\kappa}
 }
@@
 \inferrule[k-d-i]{
-	\tKindX{\tau}{\kTypeBlur}\\\\
-	\isItermX{\ibar}
+	\tKindX{\tau_1}{\kTypeBlur}\\\\
+	\tKindX{\tau_2}{\kITerm}
 }{
-	\tKindX{\tden{\ibar}{\tau}}{\kDen}
-}
-
-\inferrule[k-d-i-checked]{
-	\tKind{\Delta}{\fvalCtx}{\tau}{\kTypeBlur}\\\\
-	\isItermX{\ibar}
-}{
-	\tKind{\Delta}{\fvalCtx}{\tden{\ibar}{\tau}^\checkmark}{\kDen}
-}
-~~~~~~~
+	\tKindX{\tden{\iota}{\tau}}{\kDen}
+}
+
+%\inferrule[k-d-i-checked]{
+%	\tKind{\Delta}{\fvalCtx}{\tau}{\kTypeBlur}\\\\
+%	\isItermX{\ibar}
+%}{
+%	\tKind{\Delta}{\fvalCtx}{\tden{\ibar}{\tau}^\checkmark}{\kDen}
+%}
+%~~~~~~~
 \inferrule[k-d-e]{
@@
 }{
-	\tKindX{\titype{\sbar}}{\kIType}
+	\tKindX{\titype{\sigma}}{\kIType}
 }
 \end{mathpar}
-$\fbox{$\isItermX{\ibar}$}$
-\begin{mathpar}
-\inferrule[k-i-var]{ }{
-	\isItermX{\evar{x}}
-}
-~~~~~~
-\inferrule[k-i-fix]{
-	\isItypeX{\sbar}\\
-	\isItermX{\ibar}
-}{
-	\isItermX{\ifix{\evar{x}}{\sbar}{\ibar}}
-}
-~~~~~~
-\inferrule[k-i-transof]{
-	\tKindX{\tau}{\kDen}
-}{
-	\isItermX{\itransof{\tau}}
-}
-%\inferrule[i-lam-kinding]{
-%	\isItypeX{\sbar}\\\\
+%$\fbox{$\isItermX{\ibar}$}$
+%\begin{mathpar}
+%\inferrule[k-i-var]{ }{
+%	\isItermX{\evar{x}}
+%}
+%~~~~~~
+%\inferrule[k-i-fix]{
+%	\isItypeX{\sbar}\\
 %	\isItermX{\ibar}
 %}{
-%	\isItermX{\ilam{\evar{x}}{\sbar}{\ibar}}
+%	\isItermX{\ifix{\evar{x}}{\sbar}{\ibar}}
 %}
-
-\text{\color{gray} (omitted rules are trivially recursive, like \textsc{k-i-fix})}
-%\inferrule[iterm-dereify-kinding]{
-%	\tKindX{\tau}{\kITerm}
+%~~~~~~
+%\inferrule[k-i-transof]{
+%	\tKindX{\tau}{\kDen}
 %}{
-%	\isItermX{\iup{\tau}}
+%	\isItermX{\itransof{\tau}}
 %}
-%
-\end{mathpar}
+%%\inferrule[i-lam-kinding]{
+%%	\isItypeX{\sbar}\\\\
+%%	\isItermX{\ibar}
+%%}{
+%%	\isItermX{\ilam{\evar{x}}{\sbar}{\ibar}}
+%%}
+%
+%\text{\color{gray} (omitted rules are trivially recursive, like \textsc{k-i-fix})}
+%%\inferrule[iterm-dereify-kinding]{
+%%	\tKindX{\tau}{\kITerm}
+%%}{
+%%	\isItermX{\iup{\tau}}
+%%}
+%%
+%\end{mathpar}
 $\fbox{$\isItypeX{\sbar}$}$
@@
 
-\inferrule[k-s-repof]{
-	\tKindX{\tau}{\kTypeBlur}
-}{
-	\isItypeX{\trepof{\tau}}
-}
-
+%\inferrule[k-s-repof]{
+%	\tKindX{\tau}{\kTypeBlur}
+%}{
+%	\isItypeX{\trepof{\tau}}
+%}
+%
 \text{\color{gray} (omitted forms are trivially recursive)}
 \end{mathpar}
-\caption{\small Kinding for type-level terms.
-The kinding context $\tvarCtx$ maps type variables to kinds.
-Checked derivates (introduced by rule \textsc{k-d-i-checked}, discussed in Sec.
-\ref{repcon}) should only be constructed by the semantics (this can be enforced by a parser, so we will assume it in the metatheory).}
+\caption{\small Kinding for type-level terms. The kinding context $\tvarCtx$ maps type variables to kinds.}
 \label{tlkind}
-\vspace{-10px}
 \end{figure} 
-\begin{figure}[t]
-\small
-$\fbox{\inferrule{}{\ecompilesAX{e}{\tau}{\ibar}}}$
-\begin{mathpar}
-\inferrule[att-var]{ }{
-	\ecompilesA{\etCtxX{x}{\tau}}{\fvalCtx}{\evar{x}}{\tau}{\evar{x}}
-}
-
-\inferrule[att-lam]{
-	\tEvalX{\tau_1}{\tau_1'}\\
-%	\ddbar{\fvar{Arrow}}{\fvalCtx}{\trepof{\tau_1'}}{\sbar_1}\\
-	%\delfromtau{$\Xi_0$}{\fvalCtx}{\tau_1'}{\sabs}\\\\
-	\ecompilesA{\etCtxX{x}{\tau_1'}}{\fvalCtx}{e}{\tau_2}{\ibar}
-}{
-	\ecompilesA{\etCtx}{\fvalCtx}{\elam{\evar{x}}{\tau_1}{e}}{\ttype{Arrow}{\tpair{\tau_1'}{\tau_2}}}{\ilam{\evar{x}}{\trepof{\tau_1'}}{\ibar}}
-}
-
-\inferrule[att-op]{
-%	\tEvalX{\tauidx}{\tau_{\text{idx}}'}\\\\
-	\ecompilesA{\etCtx}{\fvalCtx}{e_1}{\tau_1}{\ibar_1}~~~~
-	\cdots~~~~
-	\ecompilesA{\etCtx}{\fvalCtx}{e_n}{\tau_n}{\ibar_n}\\
-	\fval{Tycon}{-}{-}{\theta} \in \fvalCtx\\
-	\tops{op}{\kappaidx}{i}{a}{\taudef} \in \theta\\
-	\tEvalX{\taudef~\tauidx~(\tden{\ibar_1}{\tau_1}^\checkmark :: \cdots :: \tden{\ibar_n}{\tau_n}^\checkmark :: \tnil{\kDen})}{\tinl{\kunit}{\tauD}}\\
-	\dcheck{\Gamma}{\fvalCtx}{\fvar{Arrow}, \fvar{Tycon}}{\tauD}{\tau}{\ibar}
-}{
-	\ecompilesA{\etCtx}{\fvalCtx}{\eop{Tycon}{op}{\tauidx}{\splat{e}{1}{n}}}{\tau}{\ibar}
-}
-\end{mathpar}
-\caption{\small The checked active typing judgement.
-The normalization judgement for type-level terms ($\Downarrow$) is specified in Fig.
-\ref{tleval}.
-Derivate checking is specified in Fig.
-\ref{ait}.
-Note that checked derivates can \emph{only} be constructed by the \textsc{att-op} rule (we enforce this judgementally in Sec.
-\ref{theory}).}
-\label{absatt}
-\end{figure}
-
-\section{Generic Functions, Targets and Interactive Execution}\label{targets}
-\subsection{OpenCL as an Active Library}
-The code in this section uses \verb|clx|, an example library implementing the semantics of the OpenCL programming language and extending it with some additional useful types, which we will discuss shortly.
-Ace itself has no built-in support for OpenCL.
-
-To briefly review, OpenCL provides a data-parallel SPMD programming model where developers define functions, called {\em kernels}, for execution on \emph{compute devices} like GPUs or multi-core CPUs \cite{opencl11}.
-Each thread executes the same kernel but has access to a unique index, called its \emph{global ID}.
-Kernel code is written in a variant of C99 extended with some new primitive types and operators, which we will introduce  as needed in our examples below.
-
-\subsection{Generic Functions}\label{genfn}
-%\begin{codelisting}[t]
-%\lstinputlisting[commentstyle=\color{mauve}]{listing7.py}
-%\caption{[\texttt{listing\ref{metaprogramming}.py}] Metaprogramming with Ace, showing how to construct generic functions from abstract syntax trees.}
-%\label{metaprogramming}
-%\end{codelisting}
-
-%\begin{codelisting}
-%\lstinputlisting[linebackgroundcolor={\btLstHL{5-8}}]{listing3.py}
-%\caption{[\texttt{listing\ref{map}.py}] A generic data-parallel higher-order map function targeting OpenCL.}
-%\label{map}
-%\end{codelisting}
-%\begin{codelisting}[t]
-%\lstinputlisting[commentstyle=\color{mauve}]{listing7.py}
-%\caption{[\texttt{listing\ref{metaprogramming}.py}] Metaprogramming with Ace, showing how to construct generic functions from abstract syntax trees.}
-%\label{metaprogramming}
-%\end{codelisting}
-Lines 3-4 introduce \verb|map|, an Ace function of three arguments that is governed by the \emph{active base} referred to by \verb|clx.base| and targeting the \emph{active target} referred to by \verb|clx.opencl|.
-The active target determines which language the function will compile to (here, the OpenCL kernel language) and mediates code generation.
-
-The body of this function, highlighted in grey for emphasis, does not have Python's semantics.
-Instead, it will be governed by the active base together with the active types used within it.
-No such types have been provided explicitly, however.
-Because our type system is extensible, the code inside could be meaningful for many different assignments of types to the arguments.
-We call functions awaiting types \emph{generic functions}.
-Once types have been assigned, they are called \emph{concrete functions}.
-
-Generic functions are represented at compile-time as instances of \verb|ace.GenericFn| and consist of an abstract syntax tree, an {active base} and an {active target}.
-The purpose of the \emph{decorator} on line 3 is to replace the Python function on lines 4-8 with an Ace generic function having the same syntax tree and the provided active base and active target.
-Decorators in Python are simply syntactic sugar for applying the decorator function directly to the function  being decorated \cite{python}.
-In other words, line 3 could be replaced by inserting the following  statement on line 9:
-\vspace{-0.28cm}
-\begin{verbbox}
-map = ace.fn(clx.base, clx.opencl)(map)
-\end{verbbox}
-\begin{figure}[h!]
-\centering
-\theverbbox
-\end{figure}
-\vspace{-0.28cm}
-
-The abstract syntax tree for \verb|map| is extracted using the Python standard  library packages  \verb|inspect| (to retrieve its source code) and \verb|ast| (to parse it into a syntax tree).
-
-%\subsection{Metaprogramming in Ace}
-%Generic functions can be generated directly from ASTs as well, providing Ace with support for  straightforward metaprogramming.
-Listing \ref{metaprogramming} shows how to generate two more generic functions, \verb|scale| and \verb|negate|.
-The latter is derived from the former by using a library for manipulating Python syntax trees, \verb|astx|.
-In particular, the \verb|specialize| function replaces uses of the second argument of \verb|scale| with the literal \verb|-1| (and changes the function's name), leaving a function of one argument.
-
-\subsection{Concrete Functions and Explicit Compilation}
-To compile a generic function to a particular \emph{concrete function}, a type must be provided for each argument, and typechecking and translation must then succeed.
-Listing \ref{compscript} shows how to explicitly provide type assignments to \verb|map| using the subscript operator (implemented using Python's operator overloading mechanism).
-We attempt to do so three times in Listing \ref{compscript}.
-The first, on line \ref{compscript}.7, fails due to a type error, which we handle so that the script can proceed.
-The error occurred  because the ordering of the argument types was incorrect.
-We provide a valid ordering on line \ref{compscript}.9 to generate the concrete function \verb|map_neg_f32|.
-We then provide a different type assignment to generate the concrete function \verb|map_neg_ci32|.
-Concrete functions are instances of \verb|ace.TypedFn|, consisting of an abstract syntax tree annotated with types and translations along with a reference to the original generic function.
-%The typing and translation process is mediated by the logic in the active base, types and target that have been provided, as we will describe in more detail below.
-
-To produce an output file from an Ace ``compilation script'' like \verb|listing|\texttt{\ref{compscript}}\verb|.py|, the command \verb|acec| can be invoked from the shell, as shown in Listing \ref{mapc}.
-\subsection{Types}
-\begin{codelisting}
-\lstinputlisting{listing4.py}
-\caption{[\texttt{listing\ref{compscript}.py}] The generic \texttt{map} function compiled to map the \texttt{negate} function over two  types of input.}
-\label{compscript}
-\end{codelisting}
-\begin{codelisting}
-\begin{lstlisting}[style=Bash]
-$ `acec listing3.py`
-Hello, compile-time world!
-[ace] TypeError in listing1.py (line 6, col 28): 
-      'GenericFnType(negate)' does not support [].
-[acec] listing3.cl successfully generated.
-\end{lstlisting}
-\caption{Compiling \texttt{listing\ref{compscript}.py} using the \texttt{acec} compiler.}
-\label{mapc}
-\end{codelisting}
-\begin{codelisting}
-\lstinputlisting[style=OpenCL]{listing5.cl}
-\caption{[\texttt{listing\ref{compscript}.cl}] The OpenCL file generated by Listing \ref{mapc}.}
-\label{mapout}
-\end{codelisting}
-%Lines \ref{compscript}.3-\ref{compscript}.5 construct the types assigned to the arguments of \verb|map| on lines \ref{compscript}.7-\ref{compscript}.10.
-In Ace, types are themselves values that can be manipulated at compile-time.
-This stands in contrast to other contemporary languages, where user-defined types (e.g.
-datatypes, classes, structs) are written declaratively at compile-time but cannot be constructed, inspected or passed around programmatically.
-More specifically, types are instances of a Python class that implements the \verb|ace.ActiveType| interface (see Sec.
-\ref{atypes}).
-%As Python values, types can be assigned to variables when convenient (removing the need for  facilities like \verb|typedef| in C or \verb|type| in Haskell).
-Types, like all compile-time objects derived from Ace base classes, do not have visible state and operate in a referentially transparent manner (by constructor memoization, which we do not detail here).% These types are all implemented in the \verb|clx| library imported on line 1, none are built into Ace itself.
-
-The type named \verb|T1| on line \ref{compscript}.3 corresponds to the OpenCL type \verb|global float*|: a pointer to a 32-bit floating point number stored in the compute device's global memory (one of four address spaces defined by OpenCL \cite{opencl11}).
-It is constructed by applying \verb|clx.Ptr|, which is an Ace type constructor corresponding to pointer types, to a value representing the  address space, \verb|clx.global_|, and the type being pointed to.
-That type, \verb|clx.float|, is in turn the Ace type corresponding to \verb|float| in OpenCL (which, unlike C99, is always 32 bits).
-The \verb|clx| library contains a full implementation of the OpenCL type system (including behaviors, like promotions, inherited from C99).
-Ace is \emph{unopinionated} about issues like memory safety and the wisdom of such promotions.
-We will discuss how to implement, as libraries, abstractions that are higher-level than raw pointers in Sec.
-\ref{examples}, but Ace does not prevent users from choosing a low level of abstraction or ``interesting'' semantics if the need arises (e.g.
-for compatibility with existing libraries; see the discussion in Sec.
-\ref{discussion}).
-We also note that we are being more verbose than necessary for the sake of pedagogy.
-The \verb|clx| library includes more concise shorthand for OpenCL's types: \verb|T1| is equal to \verb|clx.gp(clx.f32)|.
-%Similarly, the decorators in Listings \ref{map} and \ref{metaprogramming} could have been written \verb|clx.cl_fn|.\todo{move this back there probably}
-
-The type \verb|T2| on line \ref{compscript}.4 is a pointer to a \emph{complex integer} in global memory.
-It does not correspond direrctly to a type in OpenCL, because OpenCL does not include primitive support for complex numbers.
-Instead, it uses an active type constructor \verb|clx.Cplx|, which includes the necessary logic for typechecking operations on complex numbers and translating them to OpenCL (Sec.
-\ref{atypes}).
-This constructor is parameterized by the numeric type that should be used for the real and imaginary parts, here \verb|clx.int|, which corresponds to 32-bit OpenCL integers.
-Arithmetic operations with other complex numbers, as well as with plain numeric types (treated as if their imaginary part was zero), are supported.
-When targeting OpenCL, Ace expressions assigned type \verb|clx.Cplx(clx.int)| are compiled to OpenCL expressions of type \verb|int2|, a  \emph{vector type} of two 32-bit integers (a type that itself is not inherited from C99).
-This can be observed in several places on lines \ref{mapout}.14-\ref{mapout}.21.
-This choice is merely an implementation detail that can be kept private to \verb|clx|, however.
-An Ace value of type \verb|clx.int2| (that is, an actual OpenCL vector) \emph{cannot} be used when a \verb|clx.Cplx(clx.int)| is expected (and attempting to do so will result in a static type error): \verb|clx.Cplx| truly extends the type system, it is not a type alias.
-
-The type \verb|TF| on line \ref{compscript}.5 is extracted from the generic function \verb|negate| constructed in Listing \ref{metaprogramming}.
-Generic functions, according to Sec.
-\ref{genfn}, have not yet had a type assigned to them, so it may seem perplexing that we are nevertheless assigning a type to \verb|negate|.
-Although a conventional arrow type cannot be assigned to \verb|negate|, we can give it a \emph{singleton type}: a type that simply means ``this expression is the \emph{particular} generic function \verb|negate|''.
-This type could also have been explicitly written as \verb|ace.GenericFnType(listing2.negate)|.
-During typechecking and translation of \verb|map_neg_f32| and \verb|map_neg_ci32|, the call to \verb|f| on line \ref{map}.6 uses the type of the provided argument to compile the generic function that inhabits the singleton type of \verb|f| (\verb|negate| in both of these cases) to a concrete function.
-This is why there are two versions of \verb|negate| in the output in Listing \ref{mapout}.
-In other words, types \emph{propagate} into generic functions -- we didn't need to compile \verb|negate| explicitly.
-This also explains the error printed on line \ref{mapc}.3-\ref{mapc}.4: when this type was inadvertently assigned to the first argument \verb|input|, the indexing operation on line \ref{map}.6 resulted in an error.
-A generic function can only be \emph{statically} indexed by a list of types to turn it into a concrete function, not \emph{dynamically} indexed with a value of type \verb|clx.size_t| (the return type of the OpenCL primitive function \verb|get_global_id|).
-
-In effect, this scheme enables higher-order functions even when targeting languages, like OpenCL, that have no support for higher-order functions (OpenCL, unlike C99, does not support function pointers).
-Interestingly, because they have a singleton type, they are higher-order but not first-class functions.
-That is, the type system would prevent you from creating a heterogeneous list of generic functions.
-Concrete functions, on the other hand, can be given both a singleton type and a true function type.
-For example, \verb|listing2.negate[[clx.int]]| could be given type \verb|ace.Arrow(clx.int, clx.int)|.
-The base determines how to convert the Ace arrow type to an arrow type in the target language (e.g.
-a function pointer for C99, or an integer that indexes into a jump table constructed from knowledge of available functions of the appropriate type in OpenCL).
-
-Type assignment to generic functions is similar in some ways to template specialization in C++.
-In effect, both a template header and type parameters at call sites are being generated automatically by Ace.
-This simplifies a sophisticated feature of C++ and enables its use with other targets like OpenCL.
-%Other uses for C++ templates (and the preprocessor) are subsumed by the metaprogramming features discussed above\todo{cite template metaprogramming paper}.
-
-%\subsection{Within-Function Type Resolution}
-%\begin{codelisting}
-%\lstinputlisting{listing6.py}
-%\caption{\texttt{[listing6.py]} A function demonstrating whole-function type inference when multiple values with differing types are assigned to a single identifier, \texttt{y}.}
-%\label{inference}
-%\end{codelisting}
-%On line 5 in the generic \verb|map| function in Listing \ref{map}, the variable \verb|gid| is initialized with the result of calling the OpenCL primitive \verb|get_global_id|.
-The type for \verb|gid| is never given explicitly.
-This is a simple case of Ace's {\em within-function type resolution} strategy (we hesitate to call it \emph{type inference} because it does not, strictly speaking, take a constraint-solving approach).
-In this case, the type of \verb|gid| will resolve to \verb|size_t| because that is the return type of \verb|get_global_id| (as defined in the OpenCL specification, which the \verb|ace.OpenCL| module follows).
-The result can be observed on Lines 11 and 24 in Listing \ref{mapout}.
-%
-%Inference is not restricted within single assignments, as in the \verb|map| example, however.
-Multiple assignments to the same identifier with values of differing types, or multiple return statements, can be combined if the types in each case are compatible with one another (e.g.
-by a subtyping relation or an implicit coercion).
-In Listing \ref{inference}, the \verb|threshold_scale| function assigns different values to \verb|y| in each branch of the conditional.
-In the first branch, the value \verb|0| is an \verb|int| literal.
-However, in the second branch of the loop, the type depends on the types of both arguments, \verb|x| and \verb|scale|.
-We show two choices for these types on Lines 11 and 12.
-Type inference correctly combines these two types according to OpenCL's C99-derived rules governing numeric types (defined by the user in the \verb|OpenCL| module, as we will describe in Section \ref{att}).
-We can verify this programmatically on Lines 12 and 13.
-Note that this example would also work correctly if the assignments to \verb|y| were replaced with \verb|return| statements (in other words, the return value of a function is treated as an assignable for the purpose of type inference).
-
-\subsection{Implicit Compilation and Interactive Execution}\label{compenv}\label{backend}\label{implicit}
-%Professional end-user programmers (e.g.
-scientists and engineers \cite{professional-end-users}) today generally use dynamically-typed high-level languages like MATLAB, Python, R or Perl for tasks that are not performance-sensitive, such as small-scale data analysis and plotting \cite{nguyen2010survey}.
-For portions of their analyses where the performance overhead of dynamic type checking and automatic memory management is too high, they will typically call into code written in a statically-typed, low-level language, most commonly C or Fortran, that uses low-level parallel abstractions like pthreads and MPI \cite{4222616,basili2008understanding}.
-Unfortunately, these low-level languages and abstractions are notoriously difficult to use and automatic verification is intractable in general.\todo{integrate this}
-
-A common workflow for \emph{professional end-user programmers} (e.g.
-scientists and engineers) is to use a simple scripting language for orchestration, small-scale data analysis and visualization and call into a low-level language for performance-critical sections.
-Python is both designed for this style of use and widely adopted for such tasks \cite{sanner1999python,nguyen2010survey}.
-Developers can call into native functions using Python's foreign function interface (FFI), for example.
-A more recent trend is to generate and compile code without leaving Python, using a Python wrapper around a compiler.
-For example, \verb|weave| works with C and C++, and \verb|pycuda| and \verb|pyopencl| work with CUDA and OpenCL, respectively \cite{klockner2011pycuda}.
-\begin{codelisting}
-\lstinputlisting{listing8.py}
-\caption{[\texttt{listing\ref{py}.py}] A full OpenCL program using the \texttt{clx} Python bindings, including data transfer to and from a device and direct invocation of a generic function, \texttt{map}.}
-\label{py}
-\end{codelisting}
-The OpenCL language was designed for this workflow, exposing a retargetable compiler and data management  routines as an API, called the \emph{host API} \cite{opencl11}.
-The \verb|pyopencl| library exposes this API to Python and simplifies interoperation with \verb|numpy|, a popular package for manipulating contiguously-allocated numeric arrays in Python \cite{klockner2011pycuda}.
-
-Ace supports a refinement to this workflow, as an alternative to the \verb|acec| compiler described above, for targets that have wrappers like this available, including \verb|clx.opencl|.
-Listing \ref{py} shows an example of this workflow where the user chooses a compute device (line \ref{py}.3),  constructs a \verb|numpy| array (line \ref{py}.5), transfers it to the device (line \ref{py}.6), allocates an empty equal-sized buffer for the result of the computation (line \ref{py}.7), launches the {generic} kernel \verb|map| from Listing \ref{map} with these device arrays as well as the function \verb|negate| from Listing \ref{metaprogramming} (line \ref{py}.9) choosing a number of threads equal to the number of elements in the input array (line \ref{py}.10), and transfers the result back into main memory to check that the device computed the expected result (line \ref{py}.12).
-
-For developers experienced with the usual OpenCL or CUDA workflow, the fact that this can be accomplished in a total of 6 statements may be surprising.
-This simplicity is made possible by Ace's implicit tracking of types throughout the code.
-First, \verb|numpy| keeps track of the type, shape and order of its arrays.
-The type of \verb|input|, for example, is \verb|numpy.float64| by default, its shape is \verb|(1024,1024)| and its order is row-major by default.
-The \verb|pyopencl| library, which the \verb|clx| mechanism is built upon, uses this metadata to automatically call the underlying OpenCL host API function for transferring byte arrays to the device without requiring the user to calculate the size.
-The \verb|clx| wrapper further retains this metadata in \verb|d_input| and \verb|d_output|, the Python wrappers around the allocated device arrays.
-The Ace active type of these wrappers is an instance of \verb|clx.NPArray| parameterized by this metadata.
-This type knows how to typecheck and translate operations, like indexing with a multi-dimensional thread index, automatically.
-
-By extracting the types from the arguments, we can call the generic function \verb|map| without first requiring an explicit type assignment, like we needed when using \verb|acec| above.
-In other words, dynamic types and other metadata can propagate from Python data structures into an Ace generic function as static type information, in the same manner as it propagated \emph{between} generic functions in the previous section.
-In both cases, typechecking and translation of \verb|map| happens the first time a particular type assignment is encountered and cached for subsequent use.
-When called from Python, the generated OpenCL source code is compiled for the device we selected using the OpenCL retargetable compilation infrastructure, and cached for subsequent calls.
-
-The same program written using the OpenCL C API directly is an order of magnitude longer and significantly more difficult to comprehend.
-OpenCL does not support higher-order functions nor is there any way to write \verb|map| in a type-generic or shape-generic manner.
-If we instead use the \verb|pyopencl| library and apply the techniques described in \cite{klockner2011pycuda}, the program is still twice as large and less readable than this code.
-Both the \verb|map| and \verb|negate| functions must be explicitly specialized with the appropriate types using string manipulation techniques, and custom shapes and orders can be awkward to handle.
-Higher order functions are still not available, and must also be simulated by string manipulation.
-That approach also does not permit the use any of the language extensions that Ace enables (beyond the useful type for \verb|numpy| arrays just described; see Sec.
-\ref{examples} for more interesting possibilities).
-
-%Not shown are several additional conveniences, such as delegated kernel sizing and \verb|In| and \verb|Out| constructs that can reduce the size and improve the clarity of this code further; due to a lack of space, the reader is referred to the language documentation.
-
-\section{Related Work}\label{related}
-
-%\parindent0pt
-%
-\begin{figure*}
-\vspace{-10pt}
-\onecolumn
-\begin{longtable}{l l@{}l c@{}c c@{}c c@{}c c@{}c c@{}c}
-%\toprule%
-
-&&  &&  && {\bfseries Extensible} && {\bfseries Extensible} && {\bfseries Extensions} && {\bfseries Alternative}\\
-
-{\bfseries Approach} && {\bfseries Examples} && {\bfseries Library} && {\bfseries Syntax} && {\bfseries Type System} && {\bfseries Compositional} && {\bfseries Targets}  \\
-
-\cmidrule(l){1-1} \cmidrule(l){2-3} \cmidrule(l){4-5} \cmidrule(l){6-7} \cmidrule(l){8-9} \cmidrule(l){10-11} \cmidrule(l){12-13}
-
-\endhead
-
-Active Types && Ace && \CIRCLE && \Circle && \CIRCLE && \CIRCLE && \CIRCLE \\
-
-\myrowcolour%
-Desugaring && SugarJ \cite{erdweg2011sugarj} && \Circle && \CIRCLE && \Circle && \Circle && \Circle \\
-
-Rule Injection && Racket \cite{TypedScheme2008}, Xroma \cite{activelibraries}, mbeddr \cite{mbeddr} &&  \cite{TypedScheme2008} && \cite{mbeddr} && \CIRCLE && \Circle && \Circle \\
-
-\myrowcolour%
-Static macros /  && Scala \cite{ScalaMacros2013}, MorphJ \cite{MorphJ2011}, OJ \cite{OpenJava2000} && \Circle && \Circle && \Circle && \CIRCLE && \Circle \\
-
-\myrowcolour
-Metaprogramming &&  MetaML \cite{Sheard:1999:UMS}, Template Haskell && && && && && \\
-
-Cross-Compilation && Delite \cite{Delite2011} && \CIRCLE && \Circle && \Circle && \Circle && \CIRCLE \\
-
-\myrowcolour%
-%EDSL Frameworks && ? && \CIRCLE && \CIRCLE && \CIRCLE && \Circle && \Circle \\
-
-%Type-Specific Literals && Wyvern \cite{globaldsl13} && \Circle && \CIRCLE && \Circle && \CIRCLE && \Circle \\
-\end{longtable}
-\caption{Comparison to related approaches to language-internal extensibility.}\label{relatedtable}
-\twocolumn
-\vspace{-10pt}
-\end{figure*}
-%\subsection{Active Libraries}
-
-Some previous work has considered embedding statically-typed languages within other languages, including dynamically-typed languages.
-For example, Terra embeds a low-level statically-typed language within Lua \cite{terra}.
-Our primary example showed how to support low-level GPU programming from a high-level language.
-Other languages have supported similar workflows, e.g.
-Rust \cite{rustgpu}.
-These have not been fundamentally extensible and suffer from the ``bootstrapping'' problem described in Sec.
-\ref{intro}.
-
-Libraries containing compile-time logic have previously been called {\it active libraries} \cite{activelibraries}.
-A number of  projects, such as Blitz++, have taken advantage of C++ template metaprogramming to implement domain-specific optimizations \cite{veldhuizen2000blitz++}.
-Others have chosen custom metalanguages (e.g.
-Xroma \cite{activelibraries}, mbeddr \cite{mbeddr}).
-In Ace, we replace these brittle mini-languages with a general-purpose language, Python, and significantly expand the notion of active libraries by consideration of types, base semantics and target languages as values (in this case, objects) in the compile-time metalanguage.
-
-Generic functions are a novel strategy for {\it function polymorphism} -- defining functions that operate over more than a single type.
-In Ace, generic functions are implicitly polymorphic and can be called with arguments of {\it any type that supports the operations used by the function}.
-This is related to structural polymorphism \cite{malayeri2009structural}.
-Structural types make explicit the structure required of an argument, unlike generic functions, which are only given singleton types because the structure may depend on the semantics of an active type.
-Structural typing can be compared to the more \emph{ad hoc} approach taken by dynamically-typed languages like Python itself, sometimes called ``duck typing''.
-It is also comparable to the C++ template system, as discussed previously.
-
-{\it Operator overloading} \cite{vanWijngaarden:Mailloux:Peck:Koster:Sintzoff:Lindsey:Meertens:Fisker:acta:1975} and {\it metaobject dispatch} \cite{Kiczales91} are run-time protocols that translate operator invocations into function calls.
-The function is typically selected according to the type or value of one or more operands.
-These protocols share the notion of {\it inversion of control} with our strategy.
-However, our strategy is a {\it compile-time} protocol where the typechecking and translation semantics are implemented for different operators.
-%Our dispatch protocol was designed to be conceptually similar to Python's operator overloading protocol \cite{python}.
-
-There are several other language-internal extension mechanisms that have been described in the literature.
-These differ from our mechanism in one of the following ways, summarized in Fig.
-\ref{relatedtable}:
-\begin{itemize}
-\item Our mechanism is itself implemented as a library, rather than as a language extension. Only typed LISPs are similar.
-\item Our mechanism reuses an existing language's syntax, rather than offering syntax extension support.
-As we saw in, for example, Listing \ref{datatypest}, this can still be reasonably natural, and allows the language to benefit from a variety of existing tools.
-\item Our mechanism permits true type system extensions.
-New types are not merely aliases, nor must their rules be admissible in some base type system.% Term rewriting and macro systems do not handle types at all.
-\item As described in Sec.
-\ref{safety}, our mechanism permits compositional reasoning.
-Techniques that allow global extensions to the syntax or semantics (e.g.
-SugarJ or rule injection systems like Xroma, Typed Racket (and other typed LISPs) and mbeddr) allow extensions so much control that they can interfere.
-\item Our mechanism allows users to control the target language of compilation from within libraries. Many other mechanisms are based fundamentally on term-to-term rewriting.
-\end{itemize}
-
-When the mechanisms available in an existing language prove insufficient, researchers and domain experts often design a new language.
-A number of tools have been developed to assist with this task, including compiler generators, language workbenches and domain-specific language frameworks (cf \cite{erdweg2013state}).
-Extensible compilers can be considered a form of language framework as well due to the interoperability issues that relying on a compiler-specific extensions can introduce.
-It is difficult or impossible for these language-external approaches to achieve interoperability, because languages are not aware of each other's semantics and merging languages is not guaranteed to be sound.
-
-\section{Discussion}\label{discussion}
-
-\begin{enumerate}
-\item We address the ``chicken-and-egg'' problem by designing Ace, an extensible language  embedded within Python \cite{Politz:2013:PFM:2509136.2509536,python} entirely as a library.
-The top-level of an Ace file is a \emph{compilation script} written in Python and Python serves as Ace's \emph{type-level language}, as we will discuss.
-Ace and Python thus share a common syntax and package system, so users of Ace can directly leverage established tools, infrastructure and coding standards.
-Functions marked by a decorator as under the control of Ace are, unlike Python functions, \emph{statically} typechecked.
-This makes it possible to rule out many potential issues ahead of execution, and thus  avoid  costly dynamic checks.
-%The semantics of  define run-time behavior.
-\item  % Instead, Ace has a statically-typed semantics that can be extended by users from within libraries. 
-There is \emph{no particular type system built into Ace}.
-Instead, a type is any  object implementing the \verb|ace.Type| interface constructed by the compilation script.
-We call these \emph{active types}.
-When statically assigning a type to a compound form, e.g.
-\verb|e.x|, the Ace {compiler} delegates to the type of a subexpression, e.g.
-to the type of \verb|e|, to determine how to assign a type to the expression as a whole.
-As we will show, this {type-directed protocol}, \emph{active typing}, permits the expression of a rich variety of type systems as libraries.
-Unlike syntax-directed protocols, there cannot be ambiguities when separately defined type sytems are combined (types control non-overlapping sets of expressions by construction).
-Base cases, e.g.
-variables and certain statement forms, are handled on a {per-function basis} by a \emph{base semantics}, also a compile-time object implementing an interface, \verb|ace.Base|.
-\end{enumerate}
-% As a result, clients are able to import any combination of extensions with the confidence that link-time ambiguities cannot occur .
-
-The dynamic behavior of an Ace function is determined by translation to a \emph{target language}.
-We begin by  targeting Python, then discuss targeting OpenCL and CUDA, lower-level languages used to program many-core processors (e.g.
-GPUs).
-Active types and bases control translation by the same type-directed protocol that governs type assignment.
-We call this phase \emph{active translation}.
-A \emph{target} mediates this process and is also a user-defined compile-time object implementing an interface, \verb|ace.Target|.
-%One base or type can support multiple targets. 
-When targeting a typed language, care must be taken to ensure that well-typed Ace expressions have well-typed translations, so we integrate a technique for compositionally reasoning about compiler correctness (developed for the TIL compiler for Standard ML \cite{tarditi+:til-OLD}) into the language.
-
-TODO
-
-Static type systems are powerful tools for programming language design and implementation.
-By tracking the type of a value statically, a typechecker can verify the absence of many kinds of errors over all inputs.
-This simplifies and increases the performance of the run-time system, as errors need not be detected dynamically using tag checks and other kinds of assertions.
-
-It is legitimate to ask, however, why dynamically-typed languages are so widely-used in multiple domains.
-Although slow and difficult to reason about, these languages generally excel at satisfying the criteria of \textbf{ease-of-use}.
-More specifically, Cordy identified the principle of \emph{conciseness} as elimination of
-redundancy and the availability of reasonable defaults \cite{cordy1992hints}.
-Statically-typed languages, particularly those that professional end-users are exposed to, are often verbose, requiring explicit and often redundant type annotations on each function and variable declaration, separate header files, explicit template headers and  instantiation and other sorts of annotations.
-Dynamically-typed languages, on the other hand, avoid most of this overhead by relying on support from the run-time system.
-Ace was first conceived to explore the question: \emph{does conciseness require run-time mechanisms, or can one develop a statically-typed language with the same usability profile?}
-
-Rather than designing a new syntax, or modifying the syntax of an existing language, we chose to utilize, \emph{without modification}, the syntax of an existing language, Python.
-This choice was not arbitrary, but rather a key means by which Ace satisfies extrinsic design criteria related to familiarity and tool support.
-Researchers often dismiss the importance of syntax.
-By repurposing a well-developed syntax, they no longer need to worry about the ``trivial'' task of implementing it.
-
-Most programming languages are {\em monolithic} -- a collection of primitives are given first-class treatment by the language implementation, and users can only creatively combine them to implement  abstractions of their design.
-Although highly-expressive general-purpose mechanisms have been developed (such as object systems or functional datatypes), these may not suffice when researchers or domain experts wish to evolve aspects of the type system, exert control over the representation of data, introduce specialized run-time mechanisms, if defining an abstraction in terms of existing mechanisms is unnatural or verbose, or if custom error messages are useful.
-In these situations, it would be desirable to have the ability to modularly extend existing systems with new compile-time logic and be assured that such extensions will never interfere with one another when used in the same program.
-
-Python does not truly prevent extensions from interfering with one another because it lacks, e.g., data hiding mechanisms.
-One type could change the implementation of another by replacing its methods, for example.
-But these kinds of conflicts do not occur ``innocently'', as conflicts between two libraries in SugarJ that use similar syntax might.
-
-Ace has some limitations at the moment.
-Debuggers and other tools that rely not just on Python's syntax but also its semantics cannot be used directly, so if code generation introduces significant complexity that leads to bugs, this can be an issue.
-Debugging type errors can also be difficult when there are many nested generic functions that were implicitly given type assignments.
-Type debuggers like \verb|scalad| could help make this easier \cite{scalad}.
-We believe that active types can be used to control debugging, and plan to explore this in the future.
-We have also not yet evaluated the feasibility of implementing more advanced type systems (e.g.
-linear, dependent or flow-dependent type systems) using Ace.
-
-Not all extensions will be useful.
-Indeed, some language designers worry that offering too much flexibility to users leads to abuse (this is, for example, widely credited as the reason why Java doesn't support operator overloading).
-We do not argue with this point.
-Instead, we argue that the potential for abuse must be balanced with the possibilities made available by a vibrant ecosystem of competing statically-typed abstractions that can be developed and deployed as libraries, and thus more easily evaluated in the wild.
-With an appropriate community process, this could lead to a convergence toward stable collections of curated, high-quality collections more quickly than is possible today.
-%We also anticipate that coding guidelines and tools mandating the use of abstractions that are known to have certain desirable properties will replace designer-mandated enforcement.
-
-This paper is presented as a language design paper.
-The theoretical foundations of this work lie in type-level computation.
-We have developed a simplified, type-theoretic formalism of active typechecking and translation, where we prove the safety properties we have outlined here, as well as several additional ones that are only possible to achieve using a typed metalanguage (currently in submission at ESOP 2014).
-The work described here  extends that mechanism in several directions: Ace supports a rich syntax, makes syntax trees available to extensions, supports a pluggable backend and per-function base semantics.
-It is implemented in an existing language and supports a wider range of practical extensions.
-
-The mechanisms described here can be implemented within any language that offers some form of quotation of function ASTs and a reasonably flexible collection of syntactic forms and basic reflection facilities.
-It is particularly useful when the language supports interoperability layers with a number of target languages.
-
-%\acks
-%The author is grateful to Jonathan Aldrich, Robert Bocchino and anonymous referees for their useful suggestions.
-This work was funded by the DOE Computational Science Graduate Fellowship under grant number DE-FG02-97ER25308.
-%Acknowledgments, if needed.
-
-% We recommend abbrvnat bibliography style.
-
-\bibliographystyle{abbrv}
-
-% The bibliography should be embedded for final submission.
-
-\bibliography{../research}
-%\softraggedright
-%P. Q. Smith, and X. Y. Jones. ...reference text...
-
-\appendix
-\section{Complex Numbers}
-For example, if \verb|c| is a complex number, then \verb|c.ni| and \verb|c.i| are its non-imaginary and imaginary components, respectively.
+Some of the kinding rules for type-level terms are shown.
+%
+%\section{Complex Numbers}
+%We have implemented complex numbers (which have a logic similar to decimals, which we added late in the paper cycle). The API is somewhat dated, apologies for
+%
+%If \verb|c| is a complex number, then \verb|c.ni| and \verb|c.i| are its non-imaginary and imaginary components, respectively.
 These expressions are of the form \verb|Attribute|, so the typechecker calls \verb|type_Attribute| (line \ref{cplx}.7).
-This method receives the compilation context, \verb|context|, and the abstract syntax tree of the expression, \verb|node| and must return a type assignment for the node, or raise an \verb|ace.TypeError| if there is an error.
+%This method receives the compilation context, \verb|context|, and the abstract syntax tree of the expression, \verb|node| and must return a type assignment for the node, or raise an \verb|ace.TypeError| if there is an error.
 In this case, a type assignment is possible if the attribute name is either \verb|"ni"| or \verb|"i"|, and an error is raised otherwise (lines \ref{cplx}.8-\ref{cplx}.10).
@@
 Ace supports custom error messages.%Active types permit direct encoding of semantics and customization of error messages.
-
-Complex numbers also support binary arithmetic operations partnered with both other complex numbers and with non-complex numbers, treating them as if their imaginary component is zero.
+%
+%Complex numbers also support binary arithmetic operations partnered with both other complex numbers and with non-complex numbers, treating them as if their imaginary component is zero.
 The typechecking rules for this logic is implemented on lines \ref{cplx}.17-\ref{cplx}.29.
@@
 If supported, it applies the C99 rules for arithmetic operations to determine the resulting type (via \verb|c99_binop_t|, not shown).
-
-Finally, a complex number can be constructed inside an Ace function using Ace's special constructor form: \verb|[clx.Cplx](3,4)| represents $3+4i$, for example.
+%
+%Finally, a complex number can be constructed inside an Ace function using Ace's special constructor form: \verb|[clx.Cplx](3,4)| represents $3+4i$, for example.
 The term within the braces is evaluated at \emph{compile-time}.
@@
 It operates as expected, extracting the types of the two arguments to construct an appropriate complex number type (lines \ref{cplx}.50-\ref{cplx}.57), raising a type error if the arguments cannot be promoted to a common type according to the rules of C99 or if two arguments were not provided.% (an exercise for the reader: modify this method to also allow a single argument for when the imaginary part is 0).
-
-As seen in Listing \ref{mapout}, we are implementing complex numbers internally using OpenCL vector types, like \verb|int2|.
+%
+%As seen in Listing \ref{mapout}, we are implementing complex numbers internally using OpenCL vector types, like \verb|int2|.
 Let us look first at \verb|trans_New| on lines \ref{cplx}.54-\ref{cplx}.60, where new complex numbers are translated to vector literals by invoking \verb|target.VecLit|.
@@
 Ace supports a mode where this \emph{representational consistency} is dynamically checked during compilation (requiring that the active target know how to assign types to terms in the target language, which can be done  for our OpenCL target as of this writing).
-
-The translation methods for attributes (line \ref{cplx}.12) and binary operations  (line \ref{cplx}.31) proceed in a straightforward manner.
+%
+%The translation methods for attributes (line \ref{cplx}.12) and binary operations  (line \ref{cplx}.31) proceed in a straightforward manner.
 The context provides a method, \verb|trans|, for recursively determining the translation of subexpressions as needed.
@@
 Typechecking and translation are separated into two methods to emphasize that typechecking is not target-dependent, and to allow for more advanced uses, like type refinements and hypothetical typing judgements, that we do not describe here.
-
-\begin{codelisting}
-\lstinputlisting{listing9.py}
-\caption{\texttt{[}in \texttt{examples/clx.py]} The active type family \texttt{Ptr} implements the semantics of OpenCL pointer types.}
-\label{cplx}
-\end{codelisting}
-\section{Expressiveness}\label{examples}
-Thus far, we have focused mainly on the OpenCL target and shown examples of fairly low-level active types: those that implement OpenCL's primitives (e.g.
+%
+%\begin{codelisting}
+%\lstinputlisting{listing9.py}
+%\caption{\texttt{[}in \texttt{examples/clx.py]} The active type family \texttt{Ptr} implements the semantics of OpenCL pointer types.}
+%\label{cplx}
+%\end{codelisting}
+%\section{Expressiveness}\label{examples}
+%Thus far, we have focused mainly on the OpenCL target and shown examples of fairly low-level active types: those that implement OpenCL's primitives (e.g.
 \verb|clx.Ptr|), extend them in simple but convenient ways (e.g.
@@
 \verb|clx.NPArray|).
-Ace was first conceived to answer the question: \emph{can we build a statically-typed language with the semantics of a C but the syntax and ease-of-use of a Python?} We submit that the work described above has answered this question in the affirmative.
-
-But Ace has proven useful for more than low-level tasks like programming a GPU with OpenCL.
+%Ace was first conceived to answer the question: \emph{can we build a statically-typed language with the semantics of a C but the syntax and ease-of-use of a Python?} We submit that the work described above has answered this question in the affirmative.
+%
+%But Ace has proven useful for more than low-level tasks like programming a GPU with OpenCL.
 We now describe several interesting extensions that implement the semantics of primitives drawn from a range of different language paradigms, to justify our claim that these mechanisms are highly expressive.
-
-\subsection{Growing a Statically-Typed Python Inside an Ace}
-Ace comes with a target, base and type implementing Python itself: \verb|ace.python.python|, \verb|ace.python.base| and \verb|ace.python.dyn|.
+%
+%\subsection{Growing a Statically-Typed Python Inside an Ace}
+%Ace comes with a target, base and type implementing Python itself: \verb|ace.python.python|, \verb|ace.python.base| and \verb|ace.python.dyn|.
 These can be supplemented by additional active types and used as the foundation for writing actively-typed Python functions.
@@
 Many of the examples in the next section support this target in addition to the OpenCL target we have focused on thus far.
-
-\subsection{Recursive Labeled Sums}
+%
+\section{Recursive Labeled Sums}
+Listing 12 shows how recursive labeled sums (i.e. functional datatypes) would look.
 \begin{codelisting}
@@
 \end{codelisting}
-\begin{codelisting}
-\lstinputlisting{datatypes.py}
-\caption{\texttt{[datatypes.py]} The dynamically-typed Python code generated by running \texttt{acec datatypes\_t.py}.}
-\label{datatypes}
-\end{codelisting}
-Listing \ref{datatypest} shows an example of the use of statically-typed functional datatypes (a.k.a.
+%\begin{codelisting}
+%\lstinputlisting{datatypes.py}
+%\caption{\texttt{[datatypes.py]} The dynamically-typed Python code generated by running \texttt{acec datatypes\_t.py}.}
+%\label{datatypes}
+%\end{codelisting}
+%Listing \ref{datatypest} shows an example of the use of statically-typed functional datatypes (a.k.a.
 recursive labeled sum types) together with the Python implementation just described.
@@
 Listing \ref{datatypest} generates Listing \ref{datatypes}.
-
-Lines \ref{datatypest}.3-\ref{datatypest}.11 define a function that generates a recursive algebraic datatype representing a tree given a name and another Ace type for the data at the leaves.
+%
+%Lines \ref{datatypest}.3-\ref{datatypest}.11 define a function that generates a recursive algebraic datatype representing a tree given a name and another Ace type for the data at the leaves.
 This type implemented by the active type family \verb|fp.Datatype|.
@@
 For C-like targets, a \verb|union| type can be used (not shown).
-
-The generic function \verb|depth_gt_2| demonstrates two features.
+%
+%The generic function \verb|depth_gt_2| demonstrates two features.
 First, the base has been setup to treat the final expression in a top-level branch as the return value of the function, consistent with typical functional languages.
@@
 If the final case were omitted, for example, the algorithm would statically indicate an error, just as in statically-typed functional languages like ML.
-
-\subsection{Nested Data Parallelism}
-Functional constructs have shown promise as the basis for a nested data-parallel parallel programming model.
+%
+%%\subsection{Nested Data Parallelism}
+%%Functional constructs have shown promise as the basis for a nested data-parallel parallel programming model.
 Many powerful parallel algorithms can be specified by composing primitives like \verb|map| and \verb|reduce| on persistent data structures like lists, trees and records.
@@
 2 to implement optimizations, like fusion, on the typed syntax trees available from concrete functions, these same techniques can be implemented as an Ace library as well.
-
-\subsection{Product Types}
-Product types (types that group heterogeneous values together) like structs, unions, tuples, records and objects can all be implemented as Ace type families  parameterized by a dictionary mapping field names (indices, in the case of tuples) to types.
+%
+%\subsection{Product Types}
+%Product types (types that group heterogeneous values together) like structs, unions, tuples, records and objects can all be implemented as Ace type families  parameterized by a dictionary mapping field names (indices, in the case of tuples) to types.
 Each family differs slightly in the semantics of the operations available.
@@
 This example further demonstrates the use of argument annotations to generate concrete functions (line \ref{ooclxfppy}.9), which is a feature only available when using Python 3.3+.
-
-\subsection{Distributed Programming}
-Many distributed programming abstractions can be understood as implementing object models with complex forms of dynamic dispatch.
+%
+%\subsection{Distributed Programming}
+%Many distributed programming abstractions can be understood as implementing object models with complex forms of dynamic dispatch.
 For example, in Charm++, dynamic dispatch involves message passing over a dynamically load-balanced network \cite{kale2009charm++}.
@@
 %Erlang and other message-passing and actor-based systems can also be considered in this manner -- processes can be thought of as objects, channels as methods and messages as calls.
-
-%\subparagraph{Annotation and Extension Inference}
-%In addition to type annotations, OpenCL normally asks for additional annotations in a number of other situations.
+%
+%%\subparagraph{Annotation and Extension Inference}
+%%In addition to type annotations, OpenCL normally asks for additional annotations in a number of other situations.
 Users can annotate functions that meet certain requirements  with the \verb|__kernel| attribute, indicating that they are callable from the host.
@@
 An example of this can be seen in Listing \ref{mapout}, where the use of the \verb|double| type triggers the insertion of the \verb|cl_khr_f64| extension.
-%
-
-A number of recent languages designed for distributed programming on clusters use a partitioned global address space model, e.g.
+%%
+%
+%A number of recent languages designed for distributed programming on clusters use a partitioned global address space model, e.g.
 Chapel \cite{chapel}.
@@
 The extension mechanism can also track this information in a manner analogous to how address space information is tracked in the OpenCL example above, and target an existing portable run-time such as GASNet \cite{bonachea2002gasnet}.
-
-\subsubsection{Units of Measure}
-A number of domain-specific type systems can be implemented within Ace as well.
+%
+%\subsubsection{Units of Measure}
+%A number of domain-specific type systems can be implemented within Ace as well.
 For example, prior work has considered tracking units of measure (e.g.
@@
 Because Ace extensions can use Python, the needed logic can be implemented directly.
-
-\subsection{OOFPCLX}
-\begin{codelisting}[t]
-\lstinputlisting{ooclxfp.py}
-\caption{\texttt{[ooclxfp.py]} An example combining structs and immutable records using a prototype-based object system, cross-compiled to C99. Uses Python 3 argument annotations.}
-\label{ooclxfppy}
-\end{codelisting}
-\begin{codelisting}[t]
-\lstinputlisting[style=OpenCL]{ooclxfp.c}
-\caption{\texttt{[ooclxfp.c]} The C99 code generated by running \texttt{acec ooclxfp.py}.}
-\label{ooclxfpc}
-\end{codelisting}
+%
+%
+%\subsection{OOFPCLX}
+%\begin{codelisting}[t]
+%\lstinputlisting{ooclxfp.py}
+%\caption{\texttt{[ooclxfp.py]} An example combining structs and immutable records using a prototype-based object system, cross-compiled to C99. Uses Python 3 argument annotations.}
+%\label{ooclxfppy}
+%\end{codelisting}
+%\begin{codelisting}[t]
+%\lstinputlisting[style=OpenCL]{ooclxfp.c}
+%\caption{\texttt{[ooclxfp.c]} The C99 code generated by running \texttt{acec ooclxfp.py}.}
+%\label{ooclxfpc}
+%\end{codelisting}
 

### 2014-03-26 ace-oopsla14/datatypes_t.py

@@ -1,25 +1,24 @@
-import ace, ace.python as py, examples.fp as fp
-
-Tree = lambda name, a: fp.Datatype(name, 
+tree = lambda name, a: fp.data(name, 
   lambda tree: {
     'Empty': fp.unit,
     'Leaf': a,
     'Node': fp.tuple(tree, tree)
   })
 
-DT = Tree('DT', py.dyn)
+dyntree = tree(dyn)
 
-@ace.fn(py.Base(trailing_return=True))[[DT]]
+@py
 def depth_gt_2(x):
-  x.case({
+  {x : dyntree}
+  return x.case({
     DT.Node(DT.Node(_), _): True,
     DT.Node(_, DT.Node(_)): True,
     _: False
   })
 
-@ace.fn(py.Base(main=True))[[]]
-def __main__():
-  my_lil_tree = DT.Node(DT.Empty, DT.Empty)
-  my_big_tree = DT.Node(my_lil_tree, my_lil_tree)
+@py
+def __toplevel__():
+  my_lil_tree = dyntree.Node(dyntree.Empty, dyntree.Empty)
+  my_big_tree = dyntree.Node(my_lil_tree, my_lil_tree)
   assert not depth_gt_2(my_lil_tree)
   assert depth_gt_2(my_big_tree)

### 2014-03-26 ace-oopsla14/string-appendix.tex

@@
 for sanitizing user input. When these methods are insufficient or unavailable,
-developers often write develop custom input sanitation algorithms. In both cases,
+developers often create  custom input sanitation algorithms. In both cases,
 input sanitation techniques are ultimately implemented using the language's
@@
 XDuce typechecks XML. Like our work, XDuce relies upon some properties of
-regular expression to establish soundness and copmleteness results. Unlike
+regular expressions to establish soundness and completeness results. Unlike
 XDuce, our work is motivated by input sanitation and therefore considers arbitrary
@@
 Finally, some research languages achieve similar safety guarantees via a specialized type
-systems. The existence of such domain-specific type systems is suggestive, and 
+system. The existence of such domain-specific type systems is suggestive, and 
 we believe the simplicity of $\lcs$ demonstrates the effectiveness of
@@
 then $e \in r$. Mapping from an arbitrary $\str$ to a $\strin[r]$ requires
-defining a algorithm -- in terms of filter -- for converting a $\strin[.*]$
+defining an algorithm -- in terms of filter -- for converting a $\strin[.*]$
 into a $\strin[r]$. The static semantics of the language defines the types of
@@
 To summarize, we assume an internal language containing 
-a string typetogether with operations for string concatentation and filtering. We
+a string type together with operations for string concatentation and filtering. We
 expect closure over strings for both operations.
-Finally, recall that $\ifilter$ is only needed for dynamic casts, which may be removed without
+Finally, recall that ${\tt ifilter}$ is only needed for dynamic casts, which may be removed without
 descreasing the expressivity or even usability of the language.
@@
   \item \textbf{E-concatR, E-concatL}.
-WLOG consider E-concatR.
+Consider E-concatR.
 By induction, $e_1' : \strin{l_1}$.
@@
 Therefore, $e_1 + e_2 : \strin[r_1 + r_2]$.
+The left rule is symmetric.
   \item \textbf{E-Filterval}. We have that $\filter{\strin[r']}{e}:\strin[r\backslash r' + \emptyset]$. By inversion of T-Filter, $e:\strin[r]$.

### 2014-04-09 sanitation-plas14/sanitation-plas14.tex
Added PLAS14 template

--- 2014-03-24 ace-oopsla14/ace-oopsla14.tex
@@
 
-\title{\Ace: Growing A Statically-Typed Language Inside a Python}
+\title{A Type System for String Sanitation\\Implemented Inside a Python}
 
-\authorinfo{~}{~}{~}
-%\authorinfo{Cyrus Omar\and Nathan Fulton\and Jonathan Aldrich}
- %          {School of Computer Science\\
-  %          Carnegie Mellon University}
-   %        {\{comar, nfulton, aldrich\}@cs.cmu.edu}
+%\authorinfo{~}{~}{~}
+\authorinfo{Nathan Fulton\and Cyrus Omar\and Jonathan Aldrich}
+           {School of Computer Science\\
+            Carnegie Mellon University}
+           {\{nfulton, comar, aldrich\}@cs.cmu.edu}
 
@@
 %Evidence suggests that programmers are reluctant to adopt new languages to
-%gain access to new abstractions, even when they agree that these abstractions
-%could be valuable. This suggests a need for languages that satisfy two key
-%growth criteria: compatibility with existing programming ecosystems; and
-%internal extensibility, so that new abstractions do not require new
-%languages.
-%
-%We introduce Ace, a language compatible with the popular Python ecosystem.
-%While Python, like other similar languages, is satisfactory for simple
-%scripting, Ace is designed for more complex situations where static
-%typechecking is beneficial. Unlike most statically-typed languages, Ace's type
-%system can be extended from within, by associating compile-time functions with
-%type definitions. The compiler selectively invokes these according to a
-%type-directed protocol, avoiding interference problems. Despite these design
-%constraints, active types in Ace can express the static and dynamic semantics
-%of a range of functional, object-oriented, parallel and domain-specific
-%abstractions, all within libraries.
-Programmers are justifiably reluctant to adopt new language dialects to access stronger type systems.
-This suggests a need for a language that is \emph{compatible} with existing libraries, tools and infrastructure and that has an \emph{internally extensible type system}, so that adopting and combining type systems requires only importing libraries in the usual way, without the possibility of link-time ambiguities or safety issues.
-
-We introduce Ace, an extensible statically typed language embedded within and compatible with Python, a widely-adopted dynamically typed language.
-Python serves as Ace's type-level language.
-Rather than building in a particular type system, Ace is organized around an extensible bidirectional type system that inverts control over typechecking and translation to type-level functions according to a protocol that cannot lead to ambiguities and type safety issues when type systems are combined.
-We give a simplified calculus that describes the type theoretic foundations of this mechanism and show that, as implemented in Ace, it is flexible enough to admit practical library-based implementations of types drawn from functional, object-oriented, parallel and domain-specific languages.
-We also show how to construct safe, natural and extensible foreign function interfaces using a novel form of ``staged'' type propagation from Python into Ace.
-% can be used to enable type system extensions and demonstrates how type safety can be maintained by lifting a technique from the typed compilation literature into the language.%To show that the method scales, we give a full implementation of the OpenCL programming language as a library.
-%We briefly examine the type theoretic foundations of this approach and give modular criteria that guarantee that an extension is safe and modular.
-%Ace can be used from the shell or interactively from within Python\todo{take this sentence out?}.
-%Researchers developing languages and abstractions for high-performance computing must consider a number of design criteria, including performance, verifiability, portability and ease-of-use.
-Despite the deficiencies of legacy tools and the availability of seemingly superior options, end-users have been reluctant to adopt new language-based abstractions.
-We argue that this can be largely attributed to a failure to consider three additional criteria: continuity, extensibility\- and interoperability.
-This paper introduces Ace, a language that aims to satisfy this more comprehensive set of design criteria.
-To do so, Ace introduces several novel compile-time mechanisms, makes principled design choices, and builds upon existing standards in HPC, particularly Python and OpenCL.
-OpenCL support, rather than being built into the language, is implemented atop an extensibility mechanism that also admits abstractions drawn from other seemingly disparate paradigms.
-The core innovation underlying this and other features of Ace is a novel reification of types as first-class objects at compile-time, representing a refinement to the concept of active libraries that we call \emph{active types}.
-We validate our overall design by considering a case study of a simulation framework enabling the modular specification and efficient execution of ensembles of neural simulations across a cluster of GPUs.
+ABSTRACT HERE
 \end{abstract}
@@
 \section{Introduction}\label{intro}
-Asking programmers to import a new library is simpler than asking them to adopt a new programming language.
-Indeed, recent empirical studies underscore the difficulties of driving languages into adoption, finding that extrinsic factors like compatibility with large existing code bases,  library support, team familiarity and tool support are at least as important as intrinsic features of the language  \cite{Meyerovich:2013:EAP:2509136.2509515,chen05,nguyen2010survey}.
+INTRO HERE
 
-Unfortunately, researchers and domain experts aiming to provide potentially useful new abstractions can sometimes find it difficult to implement them as libraries, particularly when they require strengthening a language's type system.
-In these situations, abstraction providers often develop a new language or language dialect.
-Unfortunately, this \emph{language-oriented approach} \cite{journals/stp/Ward94} does not scale to large applications: using  components written in languages with different type systems can be awkward and lead to safety problems and performance overhead at interface boundaries.
-%To avoid this, it is necessary to combine the different languages and their associated tools, but this often requires significant effort and expertise (we will discuss this further in Sec.
-\ref{related-work}).
+\section{A Type System for String Sanitation}
 
-This is a problem even when a dialect introduces only a small number of new constructs.
-For example, a recent  study \cite{cave2010comparing} comparing a Java dialect, Habanero-Java (HJ), with a comparable library, \verb|java.util.concurrent|, found that the language-based abstractions in HJ were easier to use and provided useful static guarantees.
-Nevertheless, it concluded that the library-based abstractions remained more practical outside the classroom because HJ, as a distinct dialect of Java with its own type system, would be difficult to use in settings where some developers had not adopted it, but needed to interface with code that had adopted it.
-It would also be difficult to use it in combination with other abstractions also implemented as dialects of Java.
-Moreover, its tool support is more limited.
-This suggests that today, programmers and development teams cannot use  the abstractions they might prefer because they are only available bundled with languages they cannot adopt \cite{Meyerovich:2012:SDR:2414721.2414724,Meyerovich:2013:EAP:2509136.2509515}.
+\section{Implementation Inside a Python}
+\subsection{Background: Ace}
+TODO: Make a TR out of the OOPSLA submission.
+\subsection{Explicit Conversions}
+\subsection{Adding Subtyping to Ace}
+\subsection{Theory}
 
- %This issue was perhaps most succinctly expressed by a participant in a recent study by Basili et al.
-\cite{basili2008understanding} who stated ``I hate MPI, I hate C++.
-[But] if I had to choose again, I would probably choose the same.''
-%Unfortunately, researchers and domain experts (collectively, \emph{providers}) who design potentially useful new abstractions often find it difficult to implement them in terms of the general-purpose abstraction mechanisms available in existing languages.
-%Using a foreign construct can be  unnatural and, if the languages have  different type systems, can also cause safety problems.
+\section{Related Work}
 
-Internally extensible languages promise to reduce the need for new dialects by giving abstraction providers more direct control over a language's syntax and static semantics from within libraries.
-%In   an internally extensible language, programmers  can \emph{granularly} import the primitive constructs best suited to each component of their application or library.
-Providers can thus more easily develop, deploy and evaluate new abstractions in the context of existing codebases, narrowing one of the gaps between research and practice \cite{basili2008understanding}.
-%Indeed, abstractions could compete on their individual merit, rather than  by the serious adoption barriers mentioned above.
-Unfortunately, the mechanisms available today have several problems.
-First, they are themselves often available only within a dialect of an existing language and thus face a ``chicken-and-egg'' problem: a language like SugarJ \cite{erdweg2011sugarj} must overcome the same extrinsic issues as a language like HJ.
-Second, giving abstraction providers too much control over the language (by introducing an over-zealous ``solution'' to the \emph{expression problem}) can introduce type safety issues and ambiguities that only become apparent when extensions are combined, as we will discuss.
-Too little control, on the other hand, leaves it difficult to implement real-world abstractions.%It is difficult to guarantee that extensions are sound in isolation and more so to guarantee that they can be {safely} combined.
-%This requires further research before it would be appropriate to widely rely on them (which we will discuss as we go on).
-
-In this paper, we describe an extensible statically-typed language, Ace, that is implemented entirely as a library within Python, avoiding the ``chicken-and-egg problem'' and occupying what we intend to convey by example as a ``sweet spot'' in the design space (contribution 1, Sec.
-\ref{usage}).
-
-Ace, together with a core lambda calculus, serves also as a vehicle for introducing a novel extension mechanism we call \emph{active typechecking and translation}.
-We sidestep the expression problem entirely by leaving the forms of expressions fixed, sharing them with Python.
-Instead, abstraction providers extend the language by introducing new type constructors and operators using type-level functions.
-We show how, by repurposing the conventional syntactic forms found in languages like Python, structuring the language as a bidirectional type system and integrating techniques from the typed compilation literature directly into the language, we can maintain expressiveness while  guaranteeing that link-time ambiguities cannot occur and type safety is maintained (contribution 2, Secs.
-\ref{att} and \ref{theory}).
-
-%Ace can be used from the shell, producing source files that can be further compiled or executed by external means.
-Ace functions targeting a language with Python bindings (e.g.
-Python itself, as well as the others just described) can also be compiled and invoked interactively from Python.
-Ace supports a form of \emph{ad hoc} polymorphism (based on singleton types) whereby the ``dynamic type'' of a Python value can determine a static type ``just-in-time'' when Python code calls a generic Ace function.
-This represents a form of implicit \emph{staging} and we show how this can significantly streamline interactive workflows (like those in scientific computing) that alternate between Python for orchestration and exploratory tasks (e.g.
-plotting) and a statically-typed target  language via a foreign-function interface (FFI) for performance- and correctness-critical portions.
-We discuss a complete implementation of the OpenCL type system (a language for low-level data parallel programming on GPUs), show several higher-level extensions to it, and discuss deep bindings between Ace, the OpenCL run-time system and Python's standard numerics package, \verb|numpy|, all implemented within a library (contribution 3, Sec.
-\ref{targets}).%We will discuss in detail as our primary realistic case study.
-
-%The remainder of the paper is organized as follows: in Sec.
-\ref{usage}, we describe the basic structure and usage of Ace with an example that uses several interesting Ace type libraries.
-In Sec.
-\ref{att}, we detail how active typing and translation delegate control, at compile-time, to active types and bases, and discuss how this represents a form of extensibility that side-steps the expression problem.
-In Sec.
-\ref{theory}, we reduce this mechanism to  a core lambda calculus, $\lamAce$,  clarify the relationship to   type-level computation and show how type safety can be maintained.
-In Sec.
-\ref{targets}, we discuss generic functions, FFIs and staging via a complete implementation of the OpenCL type system (and extensions to it) as a library.
-
-In Sec.
-\ref{related}, we compare Ace to related work on language extensibility, term rewriting and staged compilation.
-We conclude in Sec.
-\ref{discussion} by highlighting the key features needed by a host language to support active typechecking and translation, and discussing limitations and potential future work.
-The appendix contains details omitted for concision as well as several more examples, including functional datatypes, flexible complex numbers and a previously little-explored type system for tracking string invariants using regular expressions statically (which we show being used in Listing \ref{example}, amongst several other examples).
-
-\section{Language Design and Usage}\label{usage}
-
-%
-Listing \ref{example} is an example of an Ace \emph{compilation script}.
-As promised, compilation scripts are written directly in Python.
-Ace requires no modifications to the language or features specific to the primary implementation, CPython (so Ace supports alternative implementations like Jython and PyPy).
-This choice pays immediate dividends on the first five lines: Ace's import mechanism is Python's import mechanism, so Python's build tools (e.g.
-\verb|pip|) and package repostories (e.g.
-\verb|PyPI|) are directly available for distributing Ace-based libraries.
-Note that all of the type systems that we discuss are implemented entirely as  libraries.
-%Ace is agnostic about the particular design choices that we made in our examples.
-
-%The top-level statements in an Ace file, like the \verb|print| statement on line 10, are executed to control the compile-time behavior, rather than the run-time behavior, of the program.
-%That is, Python serves as the \emph{compile-time metalanguage} (and, as we will see shortly, the \emph{type-level language}) of Ace.
-%For readers familiar with C/C++, Python can be thought of as serving a role similar to (but more general than) its preprocessor and template system (as we will see).
-
-\subsection{Types}
-\begin{codelisting}
-\begin{lstlisting}
-from examples.py import py, string
-from examples.fp import record
-from examples.oo import proto
-from examples.num import decimal
-from examples.regex import string_in
-
-print "Hello, compile-time world!"
-
-A = record['amount' : decimal[2]]
-C = record[
-  'name' : string, 
-  'account_num' : string_in[r'\d{10}'],
-  'routing_num' : string_in[r'\d{2}-\d{4}/\d{4}']]
-Transfer = proto[A, C]
-
-@py
-def log_transfer(t):
-  """Logs a transfer to the console."""
-  {t : Transfer}
-  print "Transferring %s to %s." % (
-    [string](t.amount), t.name)
-
-@py
-def __toplevel__():
-  print "Hello, run-time world!"
-  common = {name: "Annie Ace", 
-            account_num: "0000000001", 
-            routing_num: "00-0000/0001"} (C)
-  t1 = ({amount: 5.50}, common) (Transfer)
-  t2 = ({amount: 15.00}, common) (Transfer)
-  log_transfer(t1)
-  log_transfer(t2)
-
-print "Goodbye, compile-time world!"
-\end{lstlisting}
-\caption{[\texttt{listing\ref{example}.py}] An Ace compilation script.}
-\label{example}
-\end{codelisting}
-
-Types are constructed programmatically during execution of the compilation script.
-This stands in contrast to many  contemporary statically-typed languages, where types (e.g.
-datatypes, classes, structs) can only be declared.
-Put another way, Ace supports \emph{type-level computation} and Python is its type-level language (discussed further in Sec.
-\ref{theory} and Sec.
-\ref{related}).
-%The compilation script can assign shorter names to types for convenience (removing the need for  facilities like \verb|typedef| in C or \verb|type| in Haskell). 
-In our example, we see several types being constructed:
-\begin{enumerate}
-\item On line 7, we construct a functional record type with a single (immutable) field named \verb|amount| with type \verb|decimal[2]|, which classifies decimal numbers with two decimal places.
-Here, \verb|record| and \verb|decimal| are \emph{indexed type constructors}.
-We will see how they are implemented in Sec.
-\ref{att}, but note that syntactically, \verb|record| borrows Python's syntax for array slices (e.g.
-\verb|a[min:max]|).
-Note that the field name could equivalently have been written \lstinline{'a' + 'mount'}, again emphasizing that types and indices are constructed programatically by a Python script.
-\item On lines 8-11, we construct another record type.
-The field \verb|name| has type \verb|string|, defined in \verb|examples.py|, while the fields \verb|account_num| and \verb|routing_num| have more interesting types, classifying strings guaranteed statically to be in a regular language specified statically by a  regular expression pattern (written, to avoid needing to escape backslashes, using Python's \emph{raw string literals}).
-The appendix specifies the typechecking rules, based on \cite{fulton-thesis}.
-To our knowledge, this type system has not previously been implemented.
-We include it to emphasize that we aim to support specialized type systems,  not just general purpose constructs like records, numbers and strings.
-\item 
-%Prototypic inheritance  is a feature of some dynamically-typed languages (most notably, Self \cite{Ungar:Smith:oopsla:1987} and Javascript). 
-On line 12, we construct a simple \emph{prototypic object type} \cite{Lie86}.
-The type \verb|Transfer| classifies terms consisting of a \todo{what is the right term for this?}\emph{fore} of type \verb|A| and a \emph{prototype} of type \verb|C|.
-If a field cannot be found in the fore, the type system will delegate (here, statically) to the prototype.
-This makes it easy to share the values of the fields of a common prototype amongst many fores, here to allow us to describe multiple transfers differing only in amount.
-\end{enumerate}
-\subsection{Functions}
-
-Functions implementing run-time behavior (\emph{typed functions}) are distinguished by the presence of a decorator specifying their  \emph{base semantics} (here \verb|py| from \verb|examples.py|; we will discuss this further below).
-These functions are, however, still written using Python's syntax, a choice that is again valuable for extrinsic reasons: users of Ace can use a variety of  tools designed to work with Python source code  without  modification, including code highlighters (like the one used in generating this paper), editor plugins, style checkers and documentation generators.
-We will see how, with a bit of cleverness, Python's syntax can be repurposed within these functions to support a variety of static and dynamic semantics that differ from Python's.
-Ace  leverages the \verb|inspect| and \verb|ast| modules in the Python standard library to extract abstract syntax trees for typed functions \cite{python}.
-
-On lines 14-18, we see the function \verb|log_transfer|.
-As in Python, it begins with a documentation string  that informally specifies the behavior of the function.
-Before moving into the body, however, we  also write a \emph{type signature} (line 17) stating that the type of \verb|t| is \verb|Transfer|, the prototypic object type described above.
-As a result, we can assume on line 19 that \verb|t.amount| and \verb|t.name| have types \verb|decimal[2]| and \verb|string|, respectively.
-We will return to the details of \verb|print| and the form \verb|[string](t.amount)|, which performs an explicit conversion, shortly.\todo{when?}
-
-Line 17 borrows Python's syntax for dictionary literals to approximate  conventional notation for type annotations.
-%\footnote{In the rare situation where there are no type annotations and the first statement is a dictionary literal, users can use Python's \texttt{pass} keyword.} % (and following conventions for documentation strings, which also describe ``types'' at the end).
-% The ``keyword'' \verb|type| was chosen because it is the name of a built-in function in Python (and thus would not be used in the way shown for any other purpose).
- {In version 3.0 of Python, syntax for annotating arguments with arbitrary values directly was introduced \cite{pep3107}:
-% \vspace{-3px}
-\begin{lstlisting}[numbers=none]
-def log_transfer(t : Transfer): 
-  print ...
-\end{lstlisting}
-\vspace{-5px}
-These annotations were initially intended to serve only as documentation (suggesting that users of dynamically typed languages see potential in a typing discipline, if not a static one), but they were also made available as metadata for use by unspecified  future libraries.
-Ace supports both notations when the compilation script is run using Python 3.\textit{x}, but we use the more universally available notation here in view of our extrinsic goals: the Python 2.\text{x} series  remains the most widely used by a large margin as of this writing.% (though we consider cross-compilation in Sec.
-\ref{targets}).
-
-\subsection{Bidirectional Typechecking of Introductory Forms}
-On lines 21-30, we define the function \verb|__toplevel__| (the base will consider this a special name for the purposes of compilation, discussed in the Sec.
-\ref{compilation}).
-After printing a run-time greeting, we construct a \emph{value} of type \verb|C| on lines 24-26.
-Recall that \verb|C| is a record type, so we provide the names of the fields (here, without quotes because we are no longer in the type-level language) and values of the appropriate type using the syntactic form normally used for dictionary literals.
-We specify which record type the literal should be {analyzed} against by giving a \emph{literal ascription}, \verb|(C)|.
-This again repurposes existing  syntax: here function application.
-When the ``function'' is of literal form (dictionaries, tuples, lists, strings, numbers, booleans and  \verb|None|) and the argument is a type, we consider it instead an \emph{ascribed introductory form} of that type.
-We see another such form used with Python's tuple syntax on lines 27-28 to introduce terms of type \verb|Transfer| by providing the fore and prototype.
-
-%\subsection{Bidirectional Typechecking}
-The string, number and dictionary literal forms inside the outermost forms on lines 24-28 are also  introductory, but do not have an ascription.
-This is because the definition of the type \verb|C| completely determines which types they will need to have.
-As we will discuss in more detail shortly, Ace is built around a \emph{bidirectional type system} that distinguishes locations where an expression must \emph{synthesize} a type (e.g.
-to the right of bindings) from those where it can be \emph{analyzed} against a known type (e.g.
-as an argument to a function with known type) \cite{Lovas08abidirectional}.
-Literal forms are only analytic.
-
-If a literal is used in a synthetic location, the base can, however, specify a ``default ascription''.
-For example, if we had not specified the literal ascription on line 26, the base would cause the dictionary literal to be analyzed against the type \verb|dyn|, covering dynamically classified Python values.
-This would still lead to a type error because the variables used as keys are not bound in the top-level context (emphasizing that ``dynamically-typed languages'' can still have a static semantics, even if it only involves checking that top-level variables are bound).% When considered as a record, the keys were considered in a different class by the implementation (labels, rather than variables).
-
-An ascription can also evaluate statically to a {type constructor}.
-For example, we might write \verb|[1, 2] (matrix)| instead of \verb|[1, 2] (matrix[i32])| when using a base for which the default ascription for number literals is \verb|i32|.
-Because the arguments are synthetic, the type constructor can determine an appropriate index based on the types of the subexpressions.
-If we wanted a matrix of \verb|f32|s, then we would have to write either \verb|[1, 2] (matrix[f32])| or \verb|[1(f32), 2(f32)] (matrix)|, or use a base with a different default ascription.
-We will see how bidirectional typechecking of introductory forms  works in greater technical detail in Secs.
-\ref{att} and \ref{theory}, but note here that it is one of the key mechanisms for expressiveness given our constraint that we cannot add entirely new forms to the language (and indeed, doing so would begin to be quite crowded even if we could add distinct forms for matrices, vectors, lists, tuples, etc.).
-
-\subsection{External Compilation}\label{compilation} 
-\begin{codelisting}
-\begin{lstlisting}[style=Bash]
-$ `acec listing1.py`
-Hello, compile-time world!
-Goodbye, compile-time world!
-[acec] _listing1.py successfully generated.
-$ `python _listing1.py`
-Hello, run-time world!
-Transferring 5.50 to Annie Ace.
-Transferring 15.00 to Annie Ace.
-\end{lstlisting}
-\caption{Compiling \texttt{listing\ref{example}.py} using \texttt{acec}.
-Both steps can be performed at once by writing \lstinline[style=Bash]{`ace listing1.py`} (line 3 will not be printed with this command).}
-\label{external-compilation}
-\end{codelisting}
-\begin{codelisting}[t]
-\begin{lstlisting}
-import examples.num.runtime as __ace_0
-
-def print_transfer(t):
-  print "Transferring %s to %s." % (
-    __ace_0.decimal_to_str(t[0],2), t[1][0])
-
-print "Hello, run-time world!"
-common = ("Annie Ace", "0000000001", "00-0000/0000")
-t1 = ((5,50), common)
-t2 = ((15,0), common)
-log_transfer(t1)
-log_transfer(t2)
-\end{lstlisting}
-\caption{[\texttt{\_listing\ref{example}.py}] The file generated in Listing \ref{external-compilation}.}
-\label{example-out}
-\end{codelisting}
-
-To typecheck, translate and execute the code in Listing \ref{example}, we have two choices: do so externally at the shell, or perform compilation interactively from within Python.
-We will begin with the former, then return to the latter in Sec.
-\ref{targets}.% with our OpenCL examples.
-
-Listing \ref{external-compilation} shows how to invoke the \verb|acec| compiler at the shell to typecheck and translate \texttt{listing\ref{example}.py}, resulting in a file named \verb|_listing1.py|.
-This is then sent to the  Python interpreter for execution.
-Note that the {\texttt{print}} statements at the top-level of the  compilation script were evaluated during compilation only.
-These two steps can be combined by running \verb|ace listing1.py| (the intermediate file is not generated unless explicitly requested in this case).
-
-The Ace compiler is itself a Python library, and \verb|acec| is a simple Python script that invokes it, operating in two steps:
-\begin{enumerate}
-\item Evaluate the provided compilation script (\verb|listing1.py|)
-\item For any top-level bindings that are Ace functions in the final environment (instances of \verb|ace.TypedFn|, as we will discuss), it initiates active type-checking and translation (Sec.
-\ref{att}).
-If no type errors are discovered, the translations are collected (obeying order dependencies) and emitted, with file extension(s) determined by the target(s) in use, discussed in Sec.
-\ref{targets}.
-Here our target is  the default target associated with the base \verb|examples.py.py|, which emits Python 2.6 files.
-If a type error is discovered, no file is emitted and the error is displayed on the console.
-\end{enumerate}
-
-In our example, there are no type errors, so the file \verb|_listing1.py|, shown in Listing \ref{example-out},  is generated.
-This file is meant only to be executed.
-The invariants necessary to ensure that execution does not ``go wrong'' were checked statically and entities having no bearing on execution, like field names and types themselves, were erased.
-Notice that:
-\begin{enumerate}
-\item The base recognized the function name \verb|__toplevel__| as special, placing the translation of its body  at the top level of the file.
-Ace does not have any special function names itself; this is a feature of the chosen base.
-\item Records with two or more fields translated to tuples of their values (e.g.
-\verb|common| on line 8).
-If there was only a single field, like the terms of type \verb|A| inside \verb|t1| and \verb|t2|, the value was passed around unadorned.
-\item Decimals translated to pairs of integers.
-Conversion to a string happened via a helper function defined in a ``runtime'' package imported with an internal name, \verb|__ace_0|, to avoid naming conflicts.
-\item Terms of type \verb|string_in[r"..."]| translated to strings.
-Checks for membership in the specified regular language were performed entirely statically by the type system.\todo{Nathan?}\footnote{Note that there are situations (e.g.
-if a string is read in from the console) that would necessitate an initial run-time check, but no further checks in downstream functions would be necessary.
-See the appendix for details.}
-\item Prototypic objects are represented as pairs consisting of the fore and the prototype. Dispatch to the appropriate record based on the field name is static (line 5).
-\end{enumerate}
-
-% As a result, execution is essentially as fast as it could be given the target language chosen (Python). We will target an even faster language, \verb|OpenCL|, in Sec. \ref{targets}.
-
-%Because \verb|t| will be statically guaranteed to have type \verb|Transfer|,  
-\subsection{Type Errors}
-\begin{codelisting}[t]
-\begin{lstlisting}
-from listing1 import py, A, C, log_transfer
-from datetime import date
-@py
-def pay_oopsie(a):
-  {a : A}
-  print "Checking date..."
-  if date.today().day == 1:
-    common = {nome: "Oopsie Daisy",
-              account_num: None,
-              routing_num: "0-0000-0002"} (C)
-    log_transfer((common, a))
-    a.amount += 1000 
-print str(date.today().day)
-\end{lstlisting}
-\caption{[\texttt{listing\ref{oops}.py}] Lines 9-13 each have type errors.}
-\label{oops}
-\end{codelisting}
-\begin{codelisting}
-\begin{lstlisting}[style=Bash]
-$ `acec listing4.py`
-2
-[acec] TypeError in listing4.py (line 8, col 15): 
-  [record] Invalid field name: nome
-           Expected: name, account_num, routing_num
-\end{lstlisting}
-\caption{Compiling \texttt{listing\ref{oops}.py} using \texttt{acec} catches the errors statically (compilation stops at first error).}
-\label{oops-compilation}
-\end{codelisting}Listing \ref{oops} shows an example of code containing several type errors.
-Indeed, lines 9-13 each contain a type error.
-If analagous code were written in Python itself, these could only be found if the code was executed on the first day of the month (and, depending on the implementation, not all of the issues would  immediately result in run-time exceptions, possibly leading to quite subtle problems).
-Static type checking allows us to find these errors during compilation.
-Listing \ref{oops-compilation} shows the result of attempting to compile this code.
-The compilation script completes (so that functions can refer to each other mutually recursively), then the typed functions in the top-level environment are typechecked.
-The typechecker raises an exception statically at the first error, and \verb|acec| prints it as shown.
-
-\section{Active Typechecking and Translation}\label{att}
-
-In compilers and other tools for monolithic languages, there is typically a \emph{syntax-directed} protocol governing type checking, translation, pretty-printing and so on.
-For example, one might declare functional datatypes that capture all types and term forms, and then proceed by  exhaustive case analysis over terms to implement this functionality.
-The external visitor pattern for class-based object systems captures essentially the same protocol.
-In both cases, the problem is that the types and terms are defined in one place, so  providers cannot extend the language modularly.
-
-In extensible languages the focus is typically on enabling the addition of both new types and term forms.
-Adding new term forms requires then adding cases for all possible functions over terms (e.g.
-typing, translation,  code highlighting, style checking and so on).
-This is difficult.
-Indeed, this is the canonical example of the widely-studied \emph{expression problem} \cite{wadler1998expression}.
-Although there are some approaches that work for certain purposes, retaining the ability to reason compositionally about metatheoretic issues (e.g.
-type safety) and avoiding ambiguities when extensions are combined can become difficult (see Sec.
-\ref{related}).
-%Requiring that abstraction clients have the expertise to resolve these ambiguities manually is undesirable, and heading them off results in an unbounded, combinatoric explosion of situations to consider.
-
-Ace's term forms, as we have seen, are fixed by Python's grammar, so Ace sidesteps the expression problem almost entirely (and thus inherits broad tool support, as described earlier).
-The key insight is that leaving the forms fixed does not mean the semantics must also be fixed.
-Instead of taking a syntax-directed view of extensibility, we take a type-directed view: users cannot add new forms but they can add new types, which we will allow to affect the semantics of or reinterpret existing forms (e.g.
-literals, as discussed above, but also nearly every other form, as we will now discuss) in a controlled manner.
-%In Ace (but not in the theory in Sec.
-\ref{theory}), users can also repurpose existing forms flexibly within a delimited scope.
-More specifically, the key features that distinguish active typechecking and translation are: \begin{itemize}
-\item a \emph{dispatch protocol} that delegates responsibility for typechecking and translation to:
-  \begin{itemize}
-  \item a type or type constructor extracted from a subterm of a compound expression
-  \item a per-function base semantics (or simply \emph{base}) for base forms for which this is not possible  (statements, variables and unascribed literals)
-  \end{itemize}
-\item a mechanism for allowing the user to define new types, type constructors and bases from within the language (in particular, from within the type-level language) rather than fixing them ahead of time
-\end{itemize}
-\begin{codelisting}
-\begin{lstlisting}
-from listing1 import py, A
-@py
-def example(a):
-  {a : A}
-  x = a.amount
-  x = x + x
-  return {amount: x} (A)
-\end{lstlisting}
-\caption{[\texttt{listing6.py}] The example detailed in Sec. \ref{att}.}
-\label{att-example}
-\end{codelisting}
-
-To see how this works, let us trace through how Ace processes the simple example in Listing \ref{att-example}.
-
-\subsection{Base Decorators}
-Function decorators in Python (line 2) are syntactic sugar for {calling} the decorator (\verb|py|) with the decorated function (lines 3-5) as an argument, then replacing the binding (\verb|example|)  with the return value of this call.
-We could equivalently have omitted line 3 and inserted the top-level statement \verb|example = py(example)| on the line after the function definition.
-Note, however, that \verb|py| is not a function but an instance of a class, \verb|examples.py.PyBase|, that inherits from \verb|ace.Base|.
-It can be used as a decorator because \verb|ace.Base| overloads the call operator, shown in Listing \ref{base}.
-The \verb|_aceify| function (not shown) \todo{show in appendix?}operates as follows:
-\begin{enumerate}
-\item The Python standard library is used to extract an abstract syntax tree (\verb|AST|) from the function.
-\item The closure of the function, together with the globals in the module it is defined in, are extracted to form its static environment.
-\item The argument annotations are processed.
-If provided in the body, as in our example, it checks that the argument names are spelled correctly and evaluates the type-level expressions (e.g.
-\verb|A| above) in the static environment to produce a mapping of argument names to types, called the argument signature.
-If Python 3's argument annotations are used, this has  already been done for us.
-%\item Whether the function is generic or not is determined.  We defer discussion of generic functions until Sec. \ref{targets}.
-\end{enumerate}
-By the end of Listing \ref{att-example}, \verb|example| is thus an instance of \verb|ace.TypedFn|.
-It could  have been constructed directly by calling the constructor on line 2 of Listing \ref{base} (this would be inconvenient, but could be useful for metaprogramming).
-
-\subsection{Typechecking Typed Functions and Statements}
-When the Ace compiler begins typechecking a \verb|TypedFn| (when asked to by \verb|acec|, or when it is  invoked interactively as we will show in Sec. \ref{targets}), it proceeds as follows:
-\begin{enumerate}
-\item An \verb|ace.Context| is constructed with a reference to the function (Listing \ref{context}, lines 2-3).
-\item The base is asked to initialize the context.
-This can be seen on lines 5-7 of Listing \ref{pybase}, where the Python base adds an attribute tracking local bindings (initially only the arguments) and the return type to be synthesized (initially not known, indicated using \verb|None|).
-\item For each statement in the body (ignoring the documentation string and type annotation, if present), the compiler calls a method of the base, \verb|check_|$F$, where $F$ is the form of the statement (taken directly from Python's \verb|ast| package \cite{python}).
-These methods are responsible for checking that statement is well-typed (statements do not themselves have a type) and implementing any necessary effects on the context.
-
-The assignment statements on lines 5 and 6 of Listing \ref{att-example} trigger the method \verb|check_Assign| \todo{if time, add destructuring assignment to appendix}shown on lines 6-14 of Listing \ref{pybase} and the return statement on line 7 triggers the method \verb|check_Return| on lines 18-22.
-Both work similarly: if this is the first assignment to a particular identifier, or the first return statement seen, then the value must synthesize a type in the current context.
-Otherwise, it is analyzed against the previously synthesized type.
-\item The base synthesizes a type for the function as a whole via the \verb|syn_ty_FunctionDef_outer| method.
-Here, an \verb|arrow| type is synthesized (lines 26-29) based on the argument signature and the synthesized return type.
-\end{enumerate}
-
-\subsection{Bidirectionally Typechecking Expressions}
-\begin{codelisting}
-\begin{lstlisting}
-class TypedFn(object):
-  def __init__(self, base, ast, static_env, arg_sig):
-    # ...
-  # ...
-class Base(object):
-  def __call__(self, f):
-    (ast, static_env, arg_sig) = _aceify(f)
-    return TypedFn(self, ast, static_env, arg_sig)
-  # ...
-\end{lstlisting}
-\caption{A portion of the \texttt{ace} core showing how a base can be used as a decorator to construct a typed function.}
-\label{base}
-\end{codelisting}
-
-\begin{codelisting}
-\begin{lstlisting}
-class PyBase(ace.Base):
-  def init_ctx(self, ctx):
-    ctx.locals = dict(ctx.fn.arg_sig)
-    ctx.return_t = None
-
-  def check_Assign(self, ctx, s):
-    if len(s.targets) > 1:
-      raise ace.TypeError("...not supported...", s)
-    x, e = s.targets[0].id, s.value # see ast module
-	if x in ctx.locals:
-	  ctx.ana_ty(e, ctx.locals[x])
-	else:
-	  ty = ctx.syn_ty(e)
-	  ctx.locals[x] = ty
-
-  def trans_Assign(self, ctx, target, s):
-    return target.direct_translation(ctx, s)
-
-  def check_Return(self, ctx, s):
-    if ctx.return_t == None:
-      ctx.return_t = ctx.syn_ty(s.value)
-    else:
-      ctx.ana_ty(s.value, ctx.return_t)
-
-  def trans_Return(self, ctx, target, s):
-    return target.direct_translation(ctx, s)
-
-  def syn_ty_FunctionDef_outer(self, ctx, f):
-    if ctx.return_t == None:
-      ctx.return_t = unit
-    return arrow[ctx.fn.arg_sig, ctx.return_t]
-
-  def trans_FunctionDef_outer(self, ctx, target, f):
-    if f.name == "__toplevel__":
-      return target.Suite(ctx.trans(f.body))
-    else:
-      return target.direct_translation(f, ctx)
-
-  def syn_ty_Name(self, ctx, e):
-    x = e.id
-    if x in ctx.locals:
-      return ctx.locals[x]
-    elif x in ctx.fn.static_env: # (Sec. 5)
-      return self.syn_ty_lifted(x)
-    else:
-      raise ace.TypeError("...var not bound...", e)
-
-  def trans_Name(self, ctx, target, e):
-    if e.id in ctx.locals:
-      return target.direct_translation(ctx, e)
-    else: # (Sec. 5)
-      return self.trans_lifted(ctx, target, e)
-  # ...
-  default_target = PyTarget()   
-  default_Dict_ascription = dyn
-  # ...
-py = PyBase()
-\end{lstlisting}
-\caption{A portion of the base used in our examples thus far, defined in the \texttt{examples.py} package.}
-\label{pybase}
-\end{codelisting}
-
-The novel protocol for typechecking expressions that we will now describe (and formalize in Sec.
-\ref{theory}) is what enables us to extend the language with new semantics modularly while sidestepping the expression problem.
-Let us look more closely at how lines \ref{att-example}.5-\ref{att-example}.7\footnote{In this section, we will need to refer to code in Listings \ref{att-example}, \ref{pybase}, \ref{context} and \ref{record}, so we adopt this syntactic convention to refer to line numbers for concision.} are typechecked.
-
-As stated above, line \ref{att-example}.5 is an assignment statement, so the method \verb|check_Assign| on lines \ref{pybase}.6-\ref{pybase}.14 is called.
-Because the locals dictionary in \verb|ctx| does not yet know what type \verb|x| should be, the base asks to synthesize a type for the expression being bound, \verb|a.amount|, by calling the \verb|ctx.syn_ty| method.
-
-\paragraph{Active Type Synthesis} This method is defined on lines \ref{context}.5-\ref{context}.26.
-
-\subsection{Translation}
-The decorator, documentation string and argument annotations, if present, are stripped away (having already been processed).
-
-Once typechecking a method is complete, the compiler enters the translation phase, where terms in the target language are generated from Ace terms.
-Terms in the target language are generated by calling methods of the \emph{active target} governing the function being compiled.
-The translation phase operates similarly to typechecking, using the dispatch protocol to invoke methods named \verb|trans_|$X$.
-These methods have access to the context and node just as during typechecking, as well as the active target (named \verb|target| here).
-
-%When encountering a compound term (e.g.
-\verb|t.amount| in Listing \ref{example}), the compiler defers control over type  and translation to the type of a designated subexpression (e.g.
-\verb|t|) according to a fixed fixed \emph{dispatch protocol}.
-Below are examples of the choices made in Ace.\todo{full in appendix?}
-%\begin{itemize}
-%\item Responsibility over {\bf attribute access} (\texttt{e.attr}), {\bf subscripting} (\texttt{e[e1]}) and \textbf{calls} (\verb|e(e1, ..., en)|) and {\bf unary operations} (e.g.
-\verb|-e|) is handed to the type recursively assigned to \texttt{e}.
-%\item Responsibility over {\bf binary operations} (e.g.
-\verb|e1 + e2|) is first handed to the type assigned to the left operand.
-If it indicates that it cannot handle the operation, the type assigned to the right operand is handed responsibility.
-{Note that this operates like the corresponding rule in Python's \emph{dynamic} operator overloading mechanism; see Sec.
-\ref{related} for a discussion.}
-%\item Responsibility over \textbf{constructor calls} (\verb|[t](e1, ..., en)|), where \verb|t| is a \emph{compile-time Python expression} evaluating to an active type, is handed to that type.
-If \verb|t| evaluates to a \emph{family} of types, like \verb|clx.Cplx|, the active type is first generated via a class method, as discussed below.
-%%\item Responsibility over {\bf simple assignment statements}, is handed to the type of the variable on the left (which, as we will see below, is determined by the active base).
-If this type does not provide special assignment semantics, the base must handle it.%Destructuring assignment is also supported by a somewhat more complex protocol.\todo{clarify}
-%\end{itemize}
-%%
-%%\todo{To be revised from here on down}The core of Ace consists of about 1500 lines of Python code implementing its primary concepts: generic functions, concrete functions, active types, active bases and active targets.
-The latter three comprise Ace's extension mechanism.
-Extensions provide semantics to, and govern the compilation of, Ace functions, rather than logic in Ace's core.
-%Indeed, the name ``Ace'' might itself be an acronym: it is an ``active compilation environment''.
-
-\subsection{Active Types}\label{atypes}
-
-\begin{codelisting}
-\begin{lstlisting}
-class Context(object):
-  def __init__(self, fn):
-    self.fn = fn
-
-  def syn_ty(self, e):
-    if instanceof(e, ast.Name):
-      owner = self.fn.base
-      ty = owner.syn_ty_Name(self, e)
-    elif instanceof(e, ast.Attribute):
-      owner = self.syn_ty(e.value)
-      ty = owner.syn_ty_Attribute(self, e)
-    # ... other compound forms similar (cf appendix)
-    elif isinstance(e, ast.Dict):
-      return self.ana_ty(self, e, 
-        self.fn.base.default_Dict_ty)
-    # ... other literal forms similar
-    elif is_ascribed_Num(e):
-      owner = get_ascription(e)
-      if is_tycon(owner):
-        ty = owner.syn_ty_Num(self, e)
-      else:
-        return owner.ana_ty_Num(self, e)
-    # ... other ascribed literal forms similar
-    e.owner = owner
-    e.ty = ty
-    return ty
-
-  def ana_ty(self, e, ty):
-    if instanceof(e, ast.Dict):
-      ty.ana_ty_Dict(self, e)
-    # ... other literal forms similar
-    else:
-      syn_ty = self.syn_ty(e)
-      if ty != syn_ty:
-        raise TypeError("...syn/ana mismatch...", e)
-      return
-    e.owner = e.ty = ty
-
-  def trans(self, target, e):
-    if instanceof(e, ast.Name): 
-      trans = e.owner.trans_Name(self, target, e)
-    elif instanceof(e, ast.Attribute):
-      trans = e.owner.trans_Attribute(self, target, e)
-	# ... other forms similar
-	e.trans = trans
-	return trans
-\end{lstlisting}
-\caption{The context dispatches active typechecking and translation of expressions to the base or an appropriate type or type constructor.}
-\label{context}
-\end{codelisting}
-
-%Active types are the primary means for extending Ace with new abstractions.
-An active type, as mentioned previously, is an instance of a class implementing the \verb|ace.Type| interface.
-Listing \ref{record} shows an example of such a class: the \verb|clx.Cplx| class used in Listing \ref{compscript}, which implements the logic of complex numbers.
-The constructor takes as a parameter any numeric type in \verb|clx| (line \ref{cplx}.2).
-
-\subsubsection{Dispatch Protocol}
-\begin{codelisting}
-\begin{lstlisting}
-@slices_to_sig
-class record(ace.Type):
-  def __init__(self, sig):
-    _check_sig(sig)
-    self.sig = sig
-
-  def ana_ty_Dict(self, ctx, e):
-    for f, v in zip(e.keys, e.values):
-      if f.id in self.sig:
-        ctx.ana_ty(v, self.sig[f.id])
-      else:
-        raise ace.TypeError("...bad field...", f)
-    if len(self.sig) != len(e.keys):
-      raise ace.TypeError("...missing field...", e)
-
-  def trans_Dict(self, ctx, target, e):
-    if len(self.sig) == 1: # assuming non-empty
-      return ctx.trans(e.values[0])
-    else:
-      value_dict = dict(zip(e.keys, e.values))
-	  return target.Tuple(
-	    ctx.trans(target, value_dict[field])
-	    for field, ty in self.sig)
-
-  def syn_ty_Attribute(self, ctx, e): 
-    if e.attr in self.sig: 
-      return self.sig[e.attr]
-    else:
-      raise ace.TypeError("...field not found...", e)
-
-  def trans_Attribute(self, ctx, target, e): 
-    idx = idx_of(self.sig, e.attr)
-    return target.Subscript(
-      ctx.trans(target, e.value), target.Num(idx))
-\end{lstlisting}
-%
-%  def trans_type(self, target):
-%    if target.supports(examples.fp.EagerNAryProdLR):
-%      return target.EagerNAryProdLR()
-%    else if isinstance(target, examples.py.Dyn):
-%      return target.Dyn()
-\caption{The \texttt{record} type constructor.}
-\label{record}
-\end{codelisting}
-\subsubsection{Typechecking}
-When typechecking a compound expression or statement, the Ace compiler temporarily hands control to the object selected by the dispatch protocol by calling the method \verb|type_|$X$, where $X$ is  the name of the syntactic form, taken from the Python grammar \cite{python} (appended with a suffix in some cases).
-
-%Note that in each example, subexpressions are not \emph{automatically} assigned types.
-Types must be requested from the context.
-This is useful because it allows types to reinterpret subexpressions of terms they control, to supporting more complex constructs (e.g.
-pattern matching on functional datatypes, see Sec.
-\ref{datatypes}).
-
-\subsubsection{Translation}
-
-\subsection{Compositional Reasoning}\label{safety}
-Every statement and expression in an Ace function is governed by exactly one active type, determined by the dispatch protocol, or by the single active base associated with the Ace function.
-The representational consistency check ensures that compilation is successful without requiring that types that abstract over other types (e.g.
-container types) know about their internal implementation details.
-Together, this permits compositional reasoning about the semantics of Ace functions even in the presence of many extensions.
-This stands on contrast to many prior approaches to extensibility, where extensions could insert themselves into the semantics in conflicting ways, making the order of imports matter (see Sec.
-\ref{related}).
-
-\section{Theoretical Foundations}\label{theory}
-\section{Generic Functions, Targets and Interactive Execution}\label{targets}
-\subsection{OpenCL as an Active Library}
-The code in this section uses \verb|clx|, an example library implementing the semantics of the OpenCL programming language and extending it with some additional useful types, which we will discuss shortly.
-Ace itself has no built-in support for OpenCL.
-
-To briefly review, OpenCL provides a data-parallel SPMD programming model where developers define functions, called {\em kernels}, for execution on \emph{compute devices} like GPUs or multi-core CPUs \cite{opencl11}.
-Each thread executes the same kernel but has access to a unique index, called its \emph{global ID}.
-Kernel code is written in a variant of C99 extended with some new primitive types and operators, which we will introduce  as needed in our examples below.
-
-\subsection{Generic Functions}\label{genfn}
-%\begin{codelisting}[t]
-%\lstinputlisting[commentstyle=\color{mauve}]{listing7.py}
-%\caption{[\texttt{listing\ref{metaprogramming}.py}] Metaprogramming with Ace, showing how to construct generic functions from abstract syntax trees.}
-%\label{metaprogramming}
-%\end{codelisting}
-
-%\begin{codelisting}
-%\lstinputlisting[linebackgroundcolor={\btLstHL{5-8}}]{listing3.py}
-%\caption{[\texttt{listing\ref{map}.py}] A generic data-parallel higher-order map function targeting OpenCL.}
-%\label{map}
-%\end{codelisting}
-%\begin{codelisting}[t]
-%\lstinputlisting[commentstyle=\color{mauve}]{listing7.py}
-%\caption{[\texttt{listing\ref{metaprogramming}.py}] Metaprogramming with Ace, showing how to construct generic functions from abstract syntax trees.}
-%\label{metaprogramming}
-%\end{codelisting}
-Lines 3-4 introduce \verb|map|, an Ace function of three arguments that is governed by the \emph{active base} referred to by \verb|clx.base| and targeting the \emph{active target} referred to by \verb|clx.opencl|.
-The active target determines which language the function will compile to (here, the OpenCL kernel language) and mediates code generation.
-
-The body of this function, highlighted in grey for emphasis, does not have Python's semantics.
-Instead, it will be governed by the active base together with the active types used within it.
-No such types have been provided explicitly, however.
-Because our type system is extensible, the code inside could be meaningful for many different assignments of types to the arguments.
-We call functions awaiting types \emph{generic functions}.
-Once types have been assigned, they are called \emph{concrete functions}.
-
-Generic functions are represented at compile-time as instances of \verb|ace.GenericFn| and consist of an abstract syntax tree, an {active base} and an {active target}.
-The purpose of the \emph{decorator} on line 3 is to replace the Python function on lines 4-8 with an Ace generic function having the same syntax tree and the provided active base and active target.
-Decorators in Python are simply syntactic sugar for applying the decorator function directly to the function  being decorated \cite{python}.
-In other words, line 3 could be replaced by inserting the following  statement on line 9:
-\vspace{-0.28cm}
-\begin{verbbox}
-map = ace.fn(clx.base, clx.opencl)(map)
-\end{verbbox}
-\begin{figure}[h!]
-\centering
-\theverbbox
-\end{figure}
-\vspace{-0.28cm}
-
-The abstract syntax tree for \verb|map| is extracted using the Python standard  library packages  \verb|inspect| (to retrieve its source code) and \verb|ast| (to parse it into a syntax tree).
-
-\subsection{Metaprogramming in Ace}
-Generic functions can be generated directly from ASTs as well, providing Ace with support for  straightforward metaprogramming.
-Listing \ref{metaprogramming} shows how to generate two more generic functions, \verb|scale| and \verb|negate|.
-The latter is derived from the former by using a library for manipulating Python syntax trees, \verb|astx|.
-In particular, the \verb|specialize| function replaces uses of the second argument of \verb|scale| with the literal \verb|-1| (and changes the function's name), leaving a function of one argument.
-
-\subsection{Concrete Functions and Explicit Compilation}
-To compile a generic function to a particular \emph{concrete function}, a type must be provided for each argument, and typechecking and translation must then succeed.
-Listing \ref{compscript} shows how to explicitly provide type assignments to \verb|map| using the subscript operator (implemented using Python's operator overloading mechanism).
-We attempt to do so three times in Listing \ref{compscript}.
-The first, on line \ref{compscript}.7, fails due to a type error, which we handle so that the script can proceed.
-The error occurred  because the ordering of the argument types was incorrect.
-We provide a valid ordering on line \ref{compscript}.9 to generate the concrete function \verb|map_neg_f32|.
-We then provide a different type assignment to generate the concrete function \verb|map_neg_ci32|.
-Concrete functions are instances of \verb|ace.TypedFn|, consisting of an abstract syntax tree annotated with types and translations along with a reference to the original generic function.
-%The typing and translation process is mediated by the logic in the active base, types and target that have been provided, as we will describe in more detail below.
-
-To produce an output file from an Ace ``compilation script'' like \verb|listing|\texttt{\ref{compscript}}\verb|.py|, the command \verb|acec| can be invoked from the shell, as shown in Listing \ref{mapc}.
-\subsection{Types}
-\begin{codelisting}
-\lstinputlisting{listing4.py}
-\caption{[\texttt{listing\ref{compscript}.py}] The generic \texttt{map} function compiled to map the \texttt{negate} function over two  types of input.}
-\label{compscript}
-\end{codelisting}
-\begin{codelisting}
-\begin{lstlisting}[style=Bash]
-$ `acec listing3.py`
-Hello, compile-time world!
-[ace] TypeError in listing1.py (line 6, col 28): 
-      'GenericFnType(negate)' does not support [].
-[acec] listing3.cl successfully generated.
-\end{lstlisting}
-\caption{Compiling \texttt{listing\ref{compscript}.py} using the \texttt{acec} compiler.}
-\label{mapc}
-\end{codelisting}
-\begin{codelisting}
-\lstinputlisting[style=OpenCL]{listing5.cl}
-\caption{[\texttt{listing\ref{compscript}.cl}] The OpenCL file generated by Listing \ref{mapc}.}
-\label{mapout}
-\end{codelisting}
-Lines \ref{compscript}.3-\ref{compscript}.5 construct the types assigned to the arguments of \verb|map| on lines \ref{compscript}.7-\ref{compscript}.10.
-In Ace, types are themselves values that can be manipulated at compile-time.
-This stands in contrast to other contemporary languages, where user-defined types (e.g.
-datatypes, classes, structs) are written declaratively at compile-time but cannot be constructed, inspected or passed around programmatically.
-More specifically, types are instances of a Python class that implements the \verb|ace.ActiveType| interface (see Sec.
-\ref{atypes}).
-As Python values, types can be assigned to variables when convenient (removing the need for  facilities like \verb|typedef| in C or \verb|type| in Haskell).
-Types, like all compile-time objects derived from Ace base classes, do not have visible state and operate in a referentially transparent manner (by constructor memoization, which we do not detail here).% These types are all implemented in the \verb|clx| library imported on line 1, none are built into Ace itself.
-
-The type named \verb|T1| on line \ref{compscript}.3 corresponds to the OpenCL type \verb|global float*|: a pointer to a 32-bit floating point number stored in the compute device's global memory (one of four address spaces defined by OpenCL \cite{opencl11}).
-It is constructed by applying \verb|clx.Ptr|, which is an Ace type constructor corresponding to pointer types, to a value representing the  address space, \verb|clx.global_|, and the type being pointed to.
-That type, \verb|clx.float|, is in turn the Ace type corresponding to \verb|float| in OpenCL (which, unlike C99, is always 32 bits).
-The \verb|clx| library contains a full implementation of the OpenCL type system (including behaviors, like promotions, inherited from C99).
-Ace is \emph{unopinionated} about issues like memory safety and the wisdom of such promotions.
-We will discuss how to implement, as libraries, abstractions that are higher-level than raw pointers in Sec.
-\ref{examples}, but Ace does not prevent users from choosing a low level of abstraction or ``interesting'' semantics if the need arises (e.g.
-for compatibility with existing libraries; see the discussion in Sec.
-\ref{discussion}).
-We also note that we are being more verbose than necessary for the sake of pedagogy.
-The \verb|clx| library includes more concise shorthand for OpenCL's types: \verb|T1| is equal to \verb|clx.gp(clx.f32)|.
-%Similarly, the decorators in Listings \ref{map} and \ref{metaprogramming} could have been written \verb|clx.cl_fn|.\todo{move this back there probably}
-
-The type \verb|T2| on line \ref{compscript}.4 is a pointer to a \emph{complex integer} in global memory.
-It does not correspond direrctly to a type in OpenCL, because OpenCL does not include primitive support for complex numbers.
-Instead, it uses an active type constructor \verb|clx.Cplx|, which includes the necessary logic for typechecking operations on complex numbers and translating them to OpenCL (Sec.
-\ref{atypes}).
-This constructor is parameterized by the numeric type that should be used for the real and imaginary parts, here \verb|clx.int|, which corresponds to 32-bit OpenCL integers.
-Arithmetic operations with other complex numbers, as well as with plain numeric types (treated as if their imaginary part was zero), are supported.
-When targeting OpenCL, Ace expressions assigned type \verb|clx.Cplx(clx.int)| are compiled to OpenCL expressions of type \verb|int2|, a  \emph{vector type} of two 32-bit integers (a type that itself is not inherited from C99).
-This can be observed in several places on lines \ref{mapout}.14-\ref{mapout}.21.
-This choice is merely an implementation detail that can be kept private to \verb|clx|, however.
-An Ace value of type \verb|clx.int2| (that is, an actual OpenCL vector) \emph{cannot} be used when a \verb|clx.Cplx(clx.int)| is expected (and attempting to do so will result in a static type error): \verb|clx.Cplx| truly extends the type system, it is not a type alias.
-
-The type \verb|TF| on line \ref{compscript}.5 is extracted from the generic function \verb|negate| constructed in Listing \ref{metaprogramming}.
-Generic functions, according to Sec.
-\ref{genfn}, have not yet had a type assigned to them, so it may seem perplexing that we are nevertheless assigning a type to \verb|negate|.
-Although a conventional arrow type cannot be assigned to \verb|negate|, we can give it a \emph{singleton type}: a type that simply means ``this expression is the \emph{particular} generic function \verb|negate|''.
-This type could also have been explicitly written as \verb|ace.GenericFnType(listing2.negate)|.
-During typechecking and translation of \verb|map_neg_f32| and \verb|map_neg_ci32|, the call to \verb|f| on line \ref{map}.6 uses the type of the provided argument to compile the generic function that inhabits the singleton type of \verb|f| (\verb|negate| in both of these cases) to a concrete function.
-This is why there are two versions of \verb|negate| in the output in Listing \ref{mapout}.
-In other words, types \emph{propagate} into generic functions -- we didn't need to compile \verb|negate| explicitly.
-This also explains the error printed on line \ref{mapc}.3-\ref{mapc}.4: when this type was inadvertently assigned to the first argument \verb|input|, the indexing operation on line \ref{map}.6 resulted in an error.
-A generic function can only be \emph{statically} indexed by a list of types to turn it into a concrete function, not \emph{dynamically} indexed with a value of type \verb|clx.size_t| (the return type of the OpenCL primitive function \verb|get_global_id|).
-
-In effect, this scheme enables higher-order functions even when targeting languages, like OpenCL, that have no support for higher-order functions (OpenCL, unlike C99, does not support function pointers).
-Interestingly, because they have a singleton type, they are higher-order but not first-class functions.
-That is, the type system would prevent you from creating a heterogeneous list of generic functions.
-Concrete functions, on the other hand, can be given both a singleton type and a true function type.
-For example, \verb|listing2.negate[[clx.int]]| could be given type \verb|ace.Arrow(clx.int, clx.int)|.
-The base determines how to convert the Ace arrow type to an arrow type in the target language (e.g.
-a function pointer for C99, or an integer that indexes into a jump table constructed from knowledge of available functions of the appropriate type in OpenCL).
-
-Type assignment to generic functions is similar in some ways to template specialization in C++.
-In effect, both a template header and type parameters at call sites are being generated automatically by Ace.
-This simplifies a sophisticated feature of C++ and enables its use with other targets like OpenCL.
-%Other uses for C++ templates (and the preprocessor) are subsumed by the metaprogramming features discussed above\todo{cite template metaprogramming paper}.
-
-%\subsection{Within-Function Type Resolution}
-%\begin{codelisting}
-%\lstinputlisting{listing6.py}
-%\caption{\texttt{[listing6.py]} A function demonstrating whole-function type inference when multiple values with differing types are assigned to a single identifier, \texttt{y}.}
-%\label{inference}
-%\end{codelisting}
-%On line 5 in the generic \verb|map| function in Listing \ref{map}, the variable \verb|gid| is initialized with the result of calling the OpenCL primitive \verb|get_global_id|.
-The type for \verb|gid| is never given explicitly.
-This is a simple case of Ace's {\em within-function type resolution} strategy (we hesitate to call it \emph{type inference} because it does not, strictly speaking, take a constraint-solving approach).
-In this case, the type of \verb|gid| will resolve to \verb|size_t| because that is the return type of \verb|get_global_id| (as defined in the OpenCL specification, which the \verb|ace.OpenCL| module follows).
-The result can be observed on Lines 11 and 24 in Listing \ref{mapout}.
-%
-%Inference is not restricted within single assignments, as in the \verb|map| example, however.
-Multiple assignments to the same identifier with values of differing types, or multiple return statements, can be combined if the types in each case are compatible with one another (e.g.
-by a subtyping relation or an implicit coercion).
-In Listing \ref{inference}, the \verb|threshold_scale| function assigns different values to \verb|y| in each branch of the conditional.
-In the first branch, the value \verb|0| is an \verb|int| literal.
-However, in the second branch of the loop, the type depends on the types of both arguments, \verb|x| and \verb|scale|.
-We show two choices for these types on Lines 11 and 12.
-Type inference correctly combines these two types according to OpenCL's C99-derived rules governing numeric types (defined by the user in the \verb|OpenCL| module, as we will describe in Section \ref{att}).
-We can verify this programmatically on Lines 12 and 13.
-Note that this example would also work correctly if the assignments to \verb|y| were replaced with \verb|return| statements (in other words, the return value of a function is treated as an assignable for the purpose of type inference).
-
-\subsection{Implicit Compilation and Interactive Execution}\label{compenv}\label{backend}\label{implicit}
-%Professional end-user programmers (e.g.
-scientists and engineers \cite{professional-end-users}) today generally use dynamically-typed high-level languages like MATLAB, Python, R or Perl for tasks that are not performance-sensitive, such as small-scale data analysis and plotting \cite{nguyen2010survey}.
-For portions of their analyses where the performance overhead of dynamic type checking and automatic memory management is too high, they will typically call into code written in a statically-typed, low-level language, most commonly C or Fortran, that uses low-level parallel abstractions like pthreads and MPI \cite{4222616,basili2008understanding}.
-Unfortunately, these low-level languages and abstractions are notoriously difficult to use and automatic verification is intractable in general.\todo{integrate this}
-
-A common workflow for \emph{professional end-user programmers} (e.g.
-scientists and engineers) is to use a simple scripting language for orchestration, small-scale data analysis and visualization and call into a low-level language for performance-critical sections.
-Python is both designed for this style of use and widely adopted for such tasks \cite{sanner1999python,nguyen2010survey}.
-Developers can call into native functions using Python's foreign function interface (FFI), for example.
-A more recent trend is to generate and compile code without leaving Python, using a Python wrapper around a compiler.
-For example, \verb|weave| works with C and C++, and \verb|pycuda| and \verb|pyopencl| work with CUDA and OpenCL, respectively \cite{klockner2011pycuda}.
-\begin{codelisting}
-\lstinputlisting{listing8.py}
-\caption{[\texttt{listing\ref{py}.py}] A full OpenCL program using the \texttt{clx} Python bindings, including data transfer to and from a device and direct invocation of a generic function, \texttt{map}.}
-\label{py}
-\end{codelisting}
-The OpenCL language was designed for this workflow, exposing a retargetable compiler and data management  routines as an API, called the \emph{host API} \cite{opencl11}.
-The \verb|pyopencl| library exposes this API to Python and simplifies interoperation with \verb|numpy|, a popular package for manipulating contiguously-allocated numeric arrays in Python \cite{klockner2011pycuda}.
-
-Ace supports a refinement to this workflow, as an alternative to the \verb|acec| compiler described above, for targets that have wrappers like this available, including \verb|clx.opencl|.
-Listing \ref{py} shows an example of this workflow where the user chooses a compute device (line \ref{py}.3),  constructs a \verb|numpy| array (line \ref{py}.5), transfers it to the device (line \ref{py}.6), allocates an empty equal-sized buffer for the result of the computation (line \ref{py}.7), launches the {generic} kernel \verb|map| from Listing \ref{map} with these device arrays as well as the function \verb|negate| from Listing \ref{metaprogramming} (line \ref{py}.9) choosing a number of threads equal to the number of elements in the input array (line \ref{py}.10), and transfers the result back into main memory to check that the device computed the expected result (line \ref{py}.12).
-
-For developers experienced with the usual OpenCL or CUDA workflow, the fact that this can be accomplished in a total of 6 statements may be surprising.
-This simplicity is made possible by Ace's implicit tracking of types throughout the code.
-First, \verb|numpy| keeps track of the type, shape and order of its arrays.
-The type of \verb|input|, for example, is \verb|numpy.float64| by default, its shape is \verb|(1024,1024)| and its order is row-major by default.
-The \verb|pyopencl| library, which the \verb|clx| mechanism is built upon, uses this metadata to automatically call the underlying OpenCL host API function for transferring byte arrays to the device without requiring the user to calculate the size.
-The \verb|clx| wrapper further retains this metadata in \verb|d_input| and \verb|d_output|, the Python wrappers around the allocated device arrays.
-The Ace active type of these wrappers is an instance of \verb|clx.NPArray| parameterized by this metadata.
-This type knows how to typecheck and translate operations, like indexing with a multi-dimensional thread index, automatically.
-
-By extracting the types from the arguments, we can call the generic function \verb|map| without first requiring an explicit type assignment, like we needed when using \verb|acec| above.
-In other words, dynamic types and other metadata can propagate from Python data structures into an Ace generic function as static type information, in the same manner as it propagated \emph{between} generic functions in the previous section.
-In both cases, typechecking and translation of \verb|map| happens the first time a particular type assignment is encountered and cached for subsequent use.
-When called from Python, the generated OpenCL source code is compiled for the device we selected using the OpenCL retargetable compilation infrastructure, and cached for subsequent calls.
-
-The same program written using the OpenCL C API directly is an order of magnitude longer and significantly more difficult to comprehend.
-OpenCL does not support higher-order functions nor is there any way to write \verb|map| in a type-generic or shape-generic manner.
-If we instead use the \verb|pyopencl| library and apply the techniques described in \cite{klockner2011pycuda}, the program is still twice as large and less readable than this code.
-Both the \verb|map| and \verb|negate| functions must be explicitly specialized with the appropriate types using string manipulation techniques, and custom shapes and orders can be awkward to handle.
-Higher order functions are still not available, and must also be simulated by string manipulation.
-That approach also does not permit the use any of the language extensions that Ace enables (beyond the useful type for \verb|numpy| arrays just described; see Sec.
-\ref{examples} for more interesting possibilities).
-
-%Not shown are several additional conveniences, such as delegated kernel sizing and \verb|In| and \verb|Out| constructs that can reduce the size and improve the clarity of this code further; due to a lack of space, the reader is referred to the language documentation.
-
-\section{Related Work}\label{related}
-
-%\parindent0pt
-%
-\begin{figure*}
-\vspace{-10pt}
-\onecolumn
-\begin{longtable}{l l@{}l c@{}c c@{}c c@{}c c@{}c c@{}c}
-%\toprule%
-
-&&  &&  && {\bfseries Extensible} && {\bfseries Extensible} && {\bfseries Extensions} && {\bfseries Alternative}\\
-
-{\bfseries Approach} && {\bfseries Examples} && {\bfseries Library} && {\bfseries Syntax} && {\bfseries Type System} && {\bfseries Compositional} && {\bfseries Targets}  \\
-
-\cmidrule(l){1-1} \cmidrule(l){2-3} \cmidrule(l){4-5} \cmidrule(l){6-7} \cmidrule(l){8-9} \cmidrule(l){10-11} \cmidrule(l){12-13}
-
-\endhead
-
-Active Types && Ace && \CIRCLE && \Circle && \CIRCLE && \CIRCLE && \CIRCLE \\
-
-\myrowcolour%
-Desugaring && SugarJ \cite{erdweg2011sugarj} && \Circle && \CIRCLE && \Circle && \Circle && \Circle \\
-
-Rule Injection && Racket \cite{TypedScheme2008}, Xroma \cite{activelibraries}, mbeddr \cite{mbeddr} &&  \cite{TypedScheme2008} && \cite{mbeddr} && \CIRCLE && \Circle && \Circle \\
-
-\myrowcolour%
-Static macros /  && Scala \cite{ScalaMacros2013}, MorphJ \cite{MorphJ2011}, OJ \cite{OpenJava2000} && \Circle && \Circle && \Circle && \CIRCLE && \Circle \\
-
-\myrowcolour
-Metaprogramming &&  MetaML \cite{Sheard:1999:UMS}, Template Haskell && && && && && \\
-
-Cross-Compilation && Delite \cite{Delite2011} && \CIRCLE && \Circle && \Circle && \Circle && \CIRCLE \\
-
-\myrowcolour%
-%EDSL Frameworks && ? && \CIRCLE && \CIRCLE && \CIRCLE && \Circle && \Circle \\
-
-%Type-Specific Literals && Wyvern \cite{globaldsl13} && \Circle && \CIRCLE && \Circle && \CIRCLE && \Circle \\
-\end{longtable}
-\caption{Comparison to related approaches to language-internal extensibility.}\label{relatedtable}
-\twocolumn
-\vspace{-10pt}
-\end{figure*}
-%\subsection{Active Libraries}
-
-Some previous work has considered embedding statically-typed languages within other languages, including dynamically-typed languages.
-For example, Terra embeds a low-level statically-typed language within Lua \cite{terra}.
-Our primary example showed how to support low-level GPU programming from a high-level language.
-Other languages have supported similar workflows, e.g.
-Rust \cite{rustgpu}.
-These have not been fundamentally extensible and suffer from the ``bootstrapping'' problem described in Sec.
-\ref{intro}.
-
-Libraries containing compile-time logic have previously been called {\it active libraries} \cite{activelibraries}.
-A number of  projects, such as Blitz++, have taken advantage of C++ template metaprogramming to implement domain-specific optimizations \cite{veldhuizen2000blitz++}.
-Others have chosen custom metalanguages (e.g.
-Xroma \cite{activelibraries}, mbeddr \cite{mbeddr}).
-In Ace, we replace these brittle mini-languages with a general-purpose language, Python, and significantly expand the notion of active libraries by consideration of types, base semantics and target languages as values (in this case, objects) in the compile-time metalanguage.
-
-Generic functions are a novel strategy for {\it function polymorphism} -- defining functions that operate over more than a single type.
-In Ace, generic functions are implicitly polymorphic and can be called with arguments of {\it any type that supports the operations used by the function}.
-This is related to structural polymorphism \cite{malayeri2009structural}.
-Structural types make explicit the structure required of an argument, unlike generic functions, which are only given singleton types because the structure may depend on the semantics of an active type.
-Structural typing can be compared to the more \emph{ad hoc} approach taken by dynamically-typed languages like Python itself, sometimes called ``duck typing''.
-It is also comparable to the C++ template system, as discussed previously.
-
-{\it Operator overloading} \cite{vanWijngaarden:Mailloux:Peck:Koster:Sintzoff:Lindsey:Meertens:Fisker:acta:1975} and {\it metaobject dispatch} \cite{Kiczales91} are run-time protocols that translate operator invocations into function calls.
-The function is typically selected according to the type or value of one or more operands.
-These protocols share the notion of {\it inversion of control} with our strategy.
-However, our strategy is a {\it compile-time} protocol where the typechecking and translation semantics are implemented for different operators.
-%Our dispatch protocol was designed to be conceptually similar to Python's operator overloading protocol \cite{python}.
-
-There are several other language-internal extension mechanisms that have been described in the literature.
-These differ from our mechanism in one of the following ways, summarized in Fig.
-\ref{relatedtable}:
-\begin{itemize}
-\item Our mechanism is itself implemented as a library, rather than as a language extension. Only typed LISPs are similar.
-\item Our mechanism reuses an existing language's syntax, rather than offering syntax extension support.
-As we saw in, for example, Listing \ref{datatypest}, this can still be reasonably natural, and allows the language to benefit from a variety of existing tools.
-\item Our mechanism permits true type system extensions.
-New types are not merely aliases, nor must their rules be admissible in some base type system.% Term rewriting and macro systems do not handle types at all.
-\item As described in Sec.
-\ref{safety}, our mechanism permits compositional reasoning.
-Techniques that allow global extensions to the syntax or semantics (e.g.
-SugarJ or rule injection systems like Xroma, Typed Racket (and other typed LISPs) and mbeddr) allow extensions so much control that they can interfere.
-\item Our mechanism allows users to control the target language of compilation from within libraries. Many other mechanisms are based fundamentally on term-to-term rewriting.
-\end{itemize}
-
-When the mechanisms available in an existing language prove insufficient, researchers and domain experts often design a new language.
-A number of tools have been developed to assist with this task, including compiler generators, language workbenches and domain-specific language frameworks (cf \cite{erdweg2013state}).
-Extensible compilers can be considered a form of language framework as well due to the interoperability issues that relying on a compiler-specific extensions can introduce.
-It is difficult or impossible for these language-external approaches to achieve interoperability, because languages are not aware of each other's semantics and merging languages is not guaranteed to be sound.
-
-\section{Discussion}\label{discussion}
-
-\begin{enumerate}
-\item We address the ``chicken-and-egg'' problem by designing Ace, an extensible language  embedded within Python \cite{Politz:2013:PFM:2509136.2509536,python} entirely as a library.
-The top-level of an Ace file is a \emph{compilation script} written in Python and Python serves as Ace's \emph{type-level language}, as we will discuss.
-Ace and Python thus share a common syntax and package system, so users of Ace can directly leverage established tools, infrastructure and coding standards.
-Functions marked by a decorator as under the control of Ace are, unlike Python functions, \emph{statically} typechecked.
-This makes it possible to rule out many potential issues ahead of execution, and thus  avoid  costly dynamic checks.
-%The semantics of  define run-time behavior.
-\item  % Instead, Ace has a statically-typed semantics that can be extended by users from within libraries. 
-There is \emph{no particular type system built into Ace}.
-Instead, a type is any  object implementing the \verb|ace.Type| interface constructed by the compilation script.
-We call these \emph{active types}.
-When statically assigning a type to a compound form, e.g.
-\verb|e.x|, the Ace {compiler} delegates to the type of a subexpression, e.g.
-to the type of \verb|e|, to determine how to assign a type to the expression as a whole.
-As we will show, this {type-directed protocol}, \emph{active typing}, permits the expression of a rich variety of type systems as libraries.
-Unlike syntax-directed protocols, there cannot be ambiguities when separately defined type sytems are combined (types control non-overlapping sets of expressions by construction).
-Base cases, e.g.
-variables and certain statement forms, are handled on a {per-function basis} by a \emph{base semantics}, also a compile-time object implementing an interface, \verb|ace.Base|.
-\end{enumerate}
-% As a result, clients are able to import any combination of extensions with the confidence that link-time ambiguities cannot occur .
-
-The dynamic behavior of an Ace function is determined by translation to a \emph{target language}.
-We begin by  targeting Python, then discuss targeting OpenCL and CUDA, lower-level languages used to program many-core processors (e.g.
-GPUs).
-Active types and bases control translation by the same type-directed protocol that governs type assignment.
-We call this phase \emph{active translation}.
-A \emph{target} mediates this process and is also a user-defined compile-time object implementing an interface, \verb|ace.Target|.
-%One base or type can support multiple targets. 
-When targeting a typed language, care must be taken to ensure that well-typed Ace expressions have well-typed translations, so we integrate a technique for compositionally reasoning about compiler correctness (developed for the TIL compiler for Standard ML \cite{tarditi+:til-OLD}) into the language.
-
-TODO
-
-Static type systems are powerful tools for programming language design and implementation.
-By tracking the type of a value statically, a typechecker can verify the absence of many kinds of errors over all inputs.
-This simplifies and increases the performance of the run-time system, as errors need not be detected dynamically using tag checks and other kinds of assertions.
-
-It is legitimate to ask, however, why dynamically-typed languages are so widely-used in multiple domains.
-Although slow and difficult to reason about, these languages generally excel at satisfying the criteria of \textbf{ease-of-use}.
-More specifically, Cordy identified the principle of \emph{conciseness} as elimination of
-redundancy and the availability of reasonable defaults \cite{cordy1992hints}.
-Statically-typed languages, particularly those that professional end-users are exposed to, are often verbose, requiring explicit and often redundant type annotations on each function and variable declaration, separate header files, explicit template headers and  instantiation and other sorts of annotations.
-Dynamically-typed languages, on the other hand, avoid most of this overhead by relying on support from the run-time system.
-Ace was first conceived to explore the question: \emph{does conciseness require run-time mechanisms, or can one develop a statically-typed language with the same usability profile?}
-
-Rather than designing a new syntax, or modifying the syntax of an existing language, we chose to utilize, \emph{without modification}, the syntax of an existing language, Python.
-This choice was not arbitrary, but rather a key means by which Ace satisfies extrinsic design criteria related to familiarity and tool support.
-Researchers often dismiss the importance of syntax.
-By repurposing a well-developed syntax, they no longer need to worry about the ``trivial'' task of implementing it.
-
-Most programming languages are {\em monolithic} -- a collection of primitives are given first-class treatment by the language implementation, and users can only creatively combine them to implement  abstractions of their design.
-Although highly-expressive general-purpose mechanisms have been developed (such as object systems or functional datatypes), these may not suffice when researchers or domain experts wish to evolve aspects of the type system, exert control over the representation of data, introduce specialized run-time mechanisms, if defining an abstraction in terms of existing mechanisms is unnatural or verbose, or if custom error messages are useful.
-In these situations, it would be desirable to have the ability to modularly extend existing systems with new compile-time logic and be assured that such extensions will never interfere with one another when used in the same program.
-
-Python does not truly prevent extensions from interfering with one another because it lacks, e.g., data hiding mechanisms.
-One type could change the implementation of another by replacing its methods, for example.
-But these kinds of conflicts do not occur ``innocently'', as conflicts between two libraries in SugarJ that use similar syntax might.
-
-Ace has some limitations at the moment.
-Debuggers and other tools that rely not just on Python's syntax but also its semantics cannot be used directly, so if code generation introduces significant complexity that leads to bugs, this can be an issue.
-Debugging type errors can also be difficult when there are many nested generic functions that were implicitly given type assignments.
-Type debuggers like \verb|scalad| could help make this easier \cite{scalad}.
-We believe that active types can be used to control debugging, and plan to explore this in the future.
-We have also not yet evaluated the feasibility of implementing more advanced type systems (e.g.
-linear, dependent or flow-dependent type systems) using Ace.
-
-Not all extensions will be useful.
-Indeed, some language designers worry that offering too much flexibility to users leads to abuse (this is, for example, widely credited as the reason why Java doesn't support operator overloading).
-We do not argue with this point.
-Instead, we argue that the potential for abuse must be balanced with the possibilities made available by a vibrant ecosystem of competing statically-typed abstractions that can be developed and deployed as libraries, and thus more easily evaluated in the wild.
-With an appropriate community process, this could lead to a convergence toward stable collections of curated, high-quality collections more quickly than is possible today.
-%We also anticipate that coding guidelines and tools mandating the use of abstractions that are known to have certain desirable properties will replace designer-mandated enforcement.
-
-This paper is presented as a language design paper.
-The theoretical foundations of this work lie in type-level computation.
-We have developed a simplified, type-theoretic formalism of active typechecking and translation, where we prove the safety properties we have outlined here, as well as several additional ones that are only possible to achieve using a typed metalanguage (currently in submission at ESOP 2014).
-The work described here  extends that mechanism in several directions: Ace supports a rich syntax, makes syntax trees available to extensions, supports a pluggable backend and per-function base semantics.
-It is implemented in an existing language and supports a wider range of practical extensions.
-
-The mechanisms described here can be implemented within any language that offers some form of quotation of function ASTs and a reasonably flexible collection of syntactic forms and basic reflection facilities.
-It is particularly useful when the language supports interoperability layers with a number of target languages.
-
-%\acks
-%The author is grateful to Jonathan Aldrich, Robert Bocchino and anonymous referees for their useful suggestions.
-This work was funded by the DOE Computational Science Graduate Fellowship under grant number DE-FG02-97ER25308.
-%Acknowledgments, if needed.
-
-% We recommend abbrvnat bibliography style.
+\section{Discussion}
 
@@
 
-\appendix
-\section{Complex Numbers}
-For example, if \verb|c| is a complex number, then \verb|c.ni| and \verb|c.i| are its non-imaginary and imaginary components, respectively.
-These expressions are of the form \verb|Attribute|, so the typechecker calls \verb|type_Attribute| (line \ref{cplx}.7).
-This method receives the compilation context, \verb|context|, and the abstract syntax tree of the expression, \verb|node| and must return a type assignment for the node, or raise an \verb|ace.TypeError| if there is an error.
-In this case, a type assignment is possible if the attribute name is either \verb|"ni"| or \verb|"i"|, and an error is raised otherwise (lines \ref{cplx}.8-\ref{cplx}.10).
-We note that error messages are an important and sometimes overlooked facet of {ease-of-use} \cite{marceau2011measuring}.
-A common frustration with using general-purpose abstraction mechanisms to encode an abstraction is that they can produce  verbose and cryptic error messages that reflect the implementation details instead of the semantics.
-Ace supports custom error messages.%Active types permit direct encoding of semantics and customization of error messages.
-
-Complex numbers also support binary arithmetic operations partnered with both other complex numbers and with non-complex numbers, treating them as if their imaginary component is zero.
-The typechecking rules for this logic is implemented on lines \ref{cplx}.17-\ref{cplx}.29.
-Arithmetic operations are usually symmetric, so the dispatch protocol checks the types of both subexpressions for support.
-To ensure that the semantics remain deterministic in the case that both types support the binary operation, Ace asks the left first (via \verb|type_BinOp_left|), asking the right (via \verb|type_BinOp_right|) only if the left indicates an error.
-In either position, our implementation begins by recursively assigning a type to the other operand in the current context via the \verb|context.type| method (line \ref{cplx}.24).
-If supported, it applies the C99 rules for arithmetic operations to determine the resulting type (via \verb|c99_binop_t|, not shown).
-
-Finally, a complex number can be constructed inside an Ace function using Ace's special constructor form: \verb|[clx.Cplx](3,4)| represents $3+4i$, for example.
-The term within the braces is evaluated at \emph{compile-time}.
-Because \verb|clx.Cplx| evaluates not to an active type, but to a class, this form is assigned a type by handing control to the class object via the \emph{class method} \verb|type_New|.
-It operates as expected, extracting the types of the two arguments to construct an appropriate complex number type (lines \ref{cplx}.50-\ref{cplx}.57), raising a type error if the arguments cannot be promoted to a common type according to the rules of C99 or if two arguments were not provided.% (an exercise for the reader: modify this method to also allow a single argument for when the imaginary part is 0).
-
-As seen in Listing \ref{mapout}, we are implementing complex numbers internally using OpenCL vector types, like \verb|int2|.
-Let us look first at \verb|trans_New| on lines \ref{cplx}.54-\ref{cplx}.60, where new complex numbers are translated to vector literals by invoking \verb|target.VecLit|.
-This will ultimately  generate the necessary OpenCL code, as a string, to complete compilation (these strings are not directly manipulated by extensions, however, to avoid problems with, e.g.
-precedence).
-For it to be possible to reason compositionally about the correctness of compilation, all complex numbers must translate to terms in the target language that have a consistent target type.
-The \verb|trans_type| method of the \verb|ace.ActiveType| associates a type in the target language, here a vector type like \verb|int2|, with the active type.
-Ace supports a mode where this \emph{representational consistency} is dynamically checked during compilation (requiring that the active target know how to assign types to terms in the target language, which can be done  for our OpenCL target as of this writing).
-
-The translation methods for attributes (line \ref{cplx}.12) and binary operations  (line \ref{cplx}.31) proceed in a straightforward manner.
-The context provides a method, \verb|trans|, for recursively determining the translation of subexpressions as needed.
-Of note is that the translation methods can assume that typechecking succeeded.
-For example, the implementation of \verb|trans_Attribute| assumes that if \verb|node.attr| is not \verb|'ni'| then it must have been \verb|'i'| on line \ref{cplx}.14, consistent with the implementation of \verb|type_Attribute| above it.
-Typechecking and translation are separated into two methods to emphasize that typechecking is not target-dependent, and to allow for more advanced uses, like type refinements and hypothetical typing judgements, that we do not describe here.
-
-\begin{codelisting}
-\lstinputlisting{listing9.py}
-\caption{\texttt{[}in \texttt{examples/clx.py]} The active type family \texttt{Ptr} implements the semantics of OpenCL pointer types.}
-\label{cplx}
-\end{codelisting}
-\section{Expressiveness}\label{examples}
-Thus far, we have focused mainly on the OpenCL target and shown examples of fairly low-level active types: those that implement OpenCL's primitives (e.g.
-\verb|clx.Ptr|), extend them in simple but convenient ways (e.g.
-\verb|clx.Cplx|) and those that make interactive execution across language boundaries safe and convenient (e.g.
-\verb|clx.NPArray|).
-Ace was first conceived to answer the question: \emph{can we build a statically-typed language with the semantics of a C but the syntax and ease-of-use of a Python?} We submit that the work described above has answered this question in the affirmative.
-
-But Ace has proven useful for more than low-level tasks like programming a GPU with OpenCL.
-We now describe several interesting extensions that implement the semantics of primitives drawn from a range of different language paradigms, to justify our claim that these mechanisms are highly expressive.
-
-\subsection{Growing a Statically-Typed Python Inside an Ace}
-Ace comes with a target, base and type implementing Python itself: \verb|ace.python.python|, \verb|ace.python.base| and \verb|ace.python.dyn|.
-These can be supplemented by additional active types and used as the foundation for writing actively-typed Python functions.
-These functions can either be compiled ahead-of-time to an untyped Python file for execution, or be immediately executed with just-in-time compilation, just like the OpenCL examples we have shown.
-Many of the examples in the next section support this target in addition to the OpenCL target we have focused on thus far.
-
-\subsection{Recursive Labeled Sums}
-\begin{codelisting}
-\lstinputlisting{datatypes_t.py}
-\caption{\texttt{[datatypes\_t.py]} An example using statically-typed functional datatypes.}
-\label{datatypest}
-\end{codelisting}
-\begin{codelisting}
-\lstinputlisting{datatypes.py}
-\caption{\texttt{[datatypes.py]} The dynamically-typed Python code generated by running \texttt{acec datatypes\_t.py}.}
-\label{datatypes}
-\end{codelisting}
-Listing \ref{datatypest} shows an example of the use of statically-typed functional datatypes (a.k.a.
-recursive labeled sum types) together with the Python implementation just described.
-It shows two syntactic conveniences that were not mentioned previously: 1) if no target is provided to \verb|ace.fn|, the base can provide a default target (here, \verb|ace.python.python|); 2) a concrete function can be generated immediately by providing a type assignment after \verb|ace.fn| using braces.
-Listing \ref{datatypest} generates Listing \ref{datatypes}.
-
-Lines \ref{datatypest}.3-\ref{datatypest}.11 define a function that generates a recursive algebraic datatype representing a tree given a name and another Ace type for the data at the leaves.
-This type implemented by the active type family \verb|fp.Datatype|.
-A name for the datatype and the case names and types are provided programmatically on lines \ref{datatypest}.3-\ref{datatypest}.8.
-To support recursive datatypes, the case names are enclosed within a lambda term that will be passed a reference to the datatype itself.
-These lines also show two more active type families: units (the type containing just one value), and immutable pairs.
-Line \ref{datatypest}.13 calls this function with the Ace type for dynamic Python values, \verb|py.dyn|, to generate a type, aliased \verb|DT|.
-This type is implemented using class inheritance when the target is Python, as seen on lines \ref{datatypes}.1-\ref{datatypes}.13.
-For C-like targets, a \verb|union| type can be used (not shown).
-
-The generic function \verb|depth_gt_2| demonstrates two features.
-First, the base has been setup to treat the final expression in a top-level branch as the return value of the function, consistent with typical functional languages.
-Second, the \verb|case| ``method'' on line \ref{datatypest}.17 creatively reuses the syntax for dictionary literals to express a nested pattern matching construct.
-The patterns to the left of the colons are not treated as expressions (that is, a type and translation is never recursively assigned to them).
-Instead, the active type implements the standard algorithm for ensuring that the cases cover all possibilities and are not redundant \cite{pfpl}.
-If the final case were omitted, for example, the algorithm would statically indicate an error, just as in statically-typed functional languages like ML.
-
-\subsection{Nested Data Parallelism}
-Functional constructs have shown promise as the basis for a nested data-parallel parallel programming model.
-Many powerful parallel algorithms can be specified by composing primitives like \verb|map| and \verb|reduce| on persistent data structures like lists, trees and records.
-Data dependencies are directly encoded in these specifications, and automatic code generation techniques have shown promise in using this information to automatically execute these specifications on concurrent hardware and networks.
-The Copperhead programming language, for example, is based in Python as well and allows users to express functional programs over lists using common functional primitives, compiling them to CUDA code \cite{catanzaro2011copperhead}.
-By combining active types and the metaprogramming facilities described in Sec.
-2 to implement optimizations, like fusion, on the typed syntax trees available from concrete functions, these same techniques can be implemented as an Ace library as well.
-
-\subsection{Product Types}
-Product types (types that group heterogeneous values together) like structs, unions, tuples, records and objects can all be implemented as Ace type families  parameterized by a dictionary mapping field names (indices, in the case of tuples) to types.
-Each family differs slightly in the semantics of the operations available.
-Structs support mutation, for example, while records do not.
-Object systems typically introduce additional logic for calling methods, binding special variables like \verb|this|, and accessing information through an inheritance hierarchy.
-Listing \ref{ooclxfppy} shows an example of each of these.
-The prototypic object system delegates to a backing record if a field is not available in the foreground.
-This example also demonstrates cross-compilation to C99, supported for a subset of operations in \verb|clx|.
-All three abstractions are implemented as simple \verb|struct|s, as can be seen in Listing \ref{ooclxfpc}.
-This example further demonstrates the use of argument annotations to generate concrete functions (line \ref{ooclxfppy}.9), which is a feature only available when using Python 3.3+.
-
-\subsection{Distributed Programming}
-Many distributed programming abstractions can be understood as implementing object models with complex forms of dynamic dispatch.
-For example, in Charm++, dynamic dispatch involves message passing over a dynamically load-balanced network \cite{kale2009charm++}.
-While Charm++ is a sophisticated system, and we do not anticipate that implementing it as an Ace extension would be easy, it is possible to do so by targeting a system like Adaptive MPI, which exposes the Charm++ runtime system \cite{kale2009charm++}.
-%Erlang and other message-passing and actor-based systems can also be considered in this manner -- processes can be thought of as objects, channels as methods and messages as calls.
-
-%\subparagraph{Annotation and Extension Inference}
-%In addition to type annotations, OpenCL normally asks for additional annotations in a number of other situations.
-Users can annotate functions that meet certain requirements  with the \verb|__kernel| attribute, indicating that they are callable from the host.
-The \verb|OpenCL| backend can check these and add this annotation automatically.
-Several types (notably, \verb|double|) and specialized primitive functions require that an OpenCL extension be enabled via a \verb|#pragma|.
-The OpenCL backend automatically detects many of these cases as well, adding the appropriate \verb|#pragma| declaration.
-An example of this can be seen in Listing \ref{mapout}, where the use of the \verb|double| type triggers the insertion of the \verb|cl_khr_f64| extension.
-%
-
-A number of recent languages designed for distributed programming on clusters use a partitioned global address space model, e.g.
-Chapel \cite{chapel}.
-These languages provide first-class support for accessing data transparently across a massively parallel cluster.
-Their type systems track information about \emph{data locality} so that the compiler can emit more efficient code.
-The extension mechanism can also track this information in a manner analogous to how address space information is tracked in the OpenCL example above, and target an existing portable run-time such as GASNet \cite{bonachea2002gasnet}.
-
-\subsubsection{Units of Measure}
-A number of domain-specific type systems can be implemented within Ace as well.
-For example, prior work has considered tracking units of measure (e.g.
-grams) statically to validate scientific code \cite{conf/cefp/Kennedy09}.
-This cannot easily be implemented using many existing abstraction mechanisms because this information should only be maintained statically to avoid excessive run-time overhead associated with tagging/boxing, and the typechecking logic is reasonably complex.
-The Ace extension mechanism allows this information to be tracked in the type system, but not included during translation, by a mechanism similar to how address space information is tracked with pointers.
-Because Ace extensions can use Python, the needed logic can be implemented directly.
-
-\subsection{OOFPCLX}
-\begin{codelisting}[t]
-\lstinputlisting{ooclxfp.py}
-\caption{\texttt{[ooclxfp.py]} An example combining structs and immutable records using a prototype-based object system, cross-compiled to C99. Uses Python 3 argument annotations.}
-\label{ooclxfppy}
-\end{codelisting}
-\begin{codelisting}[t]
-\lstinputlisting[style=OpenCL]{ooclxfp.c}
-\caption{\texttt{[ooclxfp.c]} The C99 code generated by running \texttt{acec ooclxfp.py}.}
-\label{ooclxfpc}
-\end{codelisting}
-
 \end{document}

### 2014-04-09 sanitation-plas14/sanitation-plas14.tex
Added stuff from Appendix to the paper; messed with formatting a bit.
Begin real writing.

@@
 
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\usepackage{bussproofs}    % Gentzen-style deduction trees *with aligned sequents!*
+\usepackage{mathpartir}	% For type-settting type checking rule figures
+\usepackage{syntax}		% For type-setting formal grammars.
+
+\usepackage{hyperref}		% For links in citations
+
+\usepackage{float}			% Make figures float if [H] option is passed.
+
+\iffalse
+\usepackage{listings}		% For typesetting code listings
+\usepackage{callout}
+\usepackage{titlesec}
+\usepackage[T1]{fontenc}
+\usepackage{upquote}
+\lstset{upquote=true}
+\fi
+
+\usepackage{textcomp}		% For \textquotesingle as used in introduction
+\usepackage{color}			% for box colors, like in TAPL.
+
+\usepackage{amsmath}		% Begin Carthage default packages
+\usepackage{makeidx}
+\usepackage{amsthm}
+\usepackage{amsfonts}
+\usepackage{amssymb}
+\usepackage{graphics}
+\usepackage{enumerate}
+\usepackage{multicol}
+\usepackage{epsfig}
+\usepackage{csquotes}
+\usepackage{enumitem}
+
+\newtheorem{thm}{Theorem}                                                       
+\newtheorem{cor}[thm]{Corollary}                                                
+\newtheorem{lem}[thm]{Lemma}                                                    
+\newtheorem{prop}[thm]{Proposition}                                             
+\newtheorem{ax}[thm]{Axiom}                                                     
+\theoremstyle{definition}                                                       
+\newtheorem{defn}[thm]{Definition}                                              
+\newtheorem{exam}[thm]{Example}                                                 
+\newtheorem{rem}[thm]{Remark}
+
+%
+% For type setting inference rules with labels.
+%
+\newcommand{\inferlbl}[3]
+			{\inferrule{#3}{#2}{\textsf{\footnotesize{\sc #1}}}}
+\newcommand{\inferline}[3]
+			{\inferrule{#3}{#2} & {\textsf{\footnotesize{\sc #1}}} \\ \\}
+
+\newcommand{\Lagr}{\mathcal{L}}
+\newcommand{\strin}{{\tt string\_in}}
+\newcommand{\lang}[1]{\Lagr\{#1\}}
+\newcommand{\stru}[2]{ {\tt string\_union}(#1,#2)}
+
+\newcommand{\str}{ {\tt string} }
+\newcommand{\istr}{ {\tt istring} }
+
+\newcommand{\dconvert}[2]{ {\tt dconvert}(#1,#2) }
+\newcommand{\filter}[2]{ {\tt filter}(#1,#2) }
+\newcommand{\ifilter}[2]{ {\tt ifilter}(#1,#2) }
+
+\newcommand{\reduces}{ \Rightarrow }
+\newcommand{\ireduces}{ \leadsto }
+
+\newcommand{\val}{ \ {\tt val} }
+\newcommand{\ival}{ \ {\tt ival} }
+
+\newcommand{\istrf}[1]{`#1\textrm'} %??
+\newcommand{\strf}[1]{``#1"}
+
+%
+% Constrain the size of full-page diagrams and rule lists
+%
+%\newcommand{\pagewidth}{5in}
+%\newcommand{\rulelistwidth}{3in}
+
+%
+% Names of type systems presented in paper
+%
+\newcommand{\lcs}{\lambda_{CS}}
+
+\setlength{\grammarindent}{3em}
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
 \begin{document}
@@
 \section{Introduction}\label{intro}
-INTRO HERE
+In web applications and other settings, incorrect input sanitation often causes
+security vulnerabilities. For this reason, frameworks often provide methods
+for sanitizing user input. When these methods are insufficient or unavailable,
+developers often create  custom input sanitation algorithms. In both cases,
+input sanitation techniques are ultimately implemented using the language's
+regex library.
+
+Formally, input sanitation is the problem of ensuring that an arbitrary string
+is converted into a safe form before potentially unsafe use. For example, consider
+SQL injection attacks. To prevent such attacks, we might ensure that any string
+used as input to a query does not contain unescaped SQL command sequences.
+
+This appendix presents a type system, $\lcs$, for ensuring that input 
+sanitation algorithms are implemented correctly with respect to use site 
+specifications.
+
+Unlike frameworks provided by general purpose languages such as Haskell and
+Ruby, our type system provides a \emph{static} guarantee that input is always 
+properly sanitized before use. We achieve this by defining a typing relation
+which captures idiomatic sanitation algorithms. Type safety relies upon 
+several closure and decidability results about regular languages.
+
+XDuce typechecks XML. Like our work, XDuce relies upon some properties of
+regular expressions to establish soundness and completeness results. Unlike
+XDuce, our work is motivated by input sanitation and therefore considers arbitrary
+strings (as opposed to tree-structured XML documents). Furthermore, our static
+treatment of the ubiquitous regex filter (replace) function for regular expression is novel.
+
+Finally, some research languages achieve similar safety guarantees via a specialized type
+system. The existence of such domain-specific type systems is suggestive, and 
+we believe the simplicity of $\lcs$ demonstrates the effectiveness of
+extensibility. Instead of introducing entirely new languages for each domain,
+language extension developers may invest incrementally in obtaining static
+guarnatees which address common mistakes.
+
+An outline of this paper follows:
+
+TODO-nrf real outline!
+\begin{itemize}
+  \item In \S 2, we define the type system's static and dynamic semantics.
+  \item Section 3 recalls some classical results about regular expressions and presents a type safety proof for $\lcs$ based upon these properties.
+  \item Finally, \S 4 discusses our implemention of $\lcs$ as a type system extension  within the Ace programming language.
+\end{itemize}
 
@@
 
+The $\lcs$ language is characterized by a type of strings indexed by regular
+expressions, together with operations on such strings which correspond to common
+input sanitation patterns.
+
+This section presents the grammar and semantics of $\lcs$.
+The semantics are defined in terms of an internal language with at least strings and a regex filter function.
+These constraints are captured by the internal term valuations ($\ival$).
+The internal language does not necessarily need a regex filter function because
+any dynamic conversion is easily definable using a combination of filters and safe
+casting.
+
+%
+% Begin Grammar 
+%
+\begin{figure}
+\begin{grammar}
+<r> ::= $\epsilon$ | $.$ | $a$ | $r \cdot r$ | $r + r$ | $r*$ \hfill Regex ($a \in \Sigma$).
+
+<t> ::= 			\hfill terms: 					\\
+%$x$					\hfill variable 				\alt
+%${\tt lambda} x:t$		\hfill abstraction 			\alt
+%$t\ t$				\hfill application				\alt
+$\filter{\strin[r]}{t}$ 		\hfill filter substrings			\alt
+$[\strin[r]](t)$ \hfill safe conversion \alt
+$\dconvert{\strin[r]}{t}$ \hfill unsafe conversion
+
+<iv> ::= 			\hfill internal values:					\\
+$\istrf{s}$				\hfill internal string \alt
+${\tt ifilter}(r, \istr(s))$ \hfill internal filter
+
+<v> ::= \hfill values: \\
+$\strf{s}$ \hfill string
+
+<T> ::=				\hfill	types:					\\
+$\str$  \hfill Strings \alt
+$\strin[r]$				\hfill Regular language strings		
+%\alt
+%$T \rightarrow T$	\hfill	Functions
+
+<$\Gamma$> ::=	$\emptyset$	 | $\Gamma,x:T$ \hfill typing context
+\end{grammar}
+\caption{Syntax of $\lcs$}
+\label{fig:lcsSyntax}
+\end{figure}
+
+The $\lcs$ language gives static semantics for common regular expression library
+functions. In this treatment, we include concatenation and filtering.
+The ${\tt filter}$ function removes all instances of a regular expression in a string,
+while concatenation ($+$) concatenates two strings.
+
+\subsection{Typing}
+
+The $\strin[r]$ type is parameterized by regular expressions; if $e:\strin[r]$,
+then $e \in r$. Mapping from an arbitrary $\str$ to a $\strin[r]$ requires
+defining an algorithm -- in terms of filter -- for converting a $\strin[.*]$
+into a $\strin[r]$. The static semantics of the language defines the types of
+operations on regular expressions in terms of well-understood properties about
+regular lanugages; we recall these properties in section 3.
+
+\subsection{Dynamics}
+
+There are two evaluation judgements: $e:T \reduces e'$ and $e:T \ireduces i$.
+The $\reduces$ relation is between $\lcs$ expressions, while the $\ireduces$
+relation is a mapping from $\lcs$ expressions into internal language expressions
+$i$ such that $i \ival$.
+
+Safety of the evaluation relation depends upon an injective mapping from $\lcs$ types info
+internal language types. This relation, $h$, is defined below.
+
+\subsection{Type Safety}
+
+The type safety proof relies upon some assumptions about the type system and
+dynamics of the internal language, as well as some properties of regular
+languages.
+
+There must exist a translation from $\lcs$ types to the types of the internal
+language. For the remainder of this paper, we call the type translation function
+$h$.
+
+\begin{defn}[Type Translation Function $h$]
+  The type translation function $h : Type \rightarrow IType$ is defined as follows:
+  \begin{itemize}
+    \item $\forall r. h(\strin[r]) = \istr$
+    \item $h(\str) = \istr$
+  \end{itemize}
+\end{defn}
+
+Additionally, we assume that the internal language contains an implementation of strings,
+together with operations for concatenation and filtering by regular expression.
+
+\begin{defn}[Types of internal values]
+  Let $\istrf{s}$ range over string literals and $r$ over regular expressions.
+  Internal values are typed as follows:
+\begin{itemize}
+  \item If $e=\istrf{s}$ then $e:\istr$.
+  \item If $e=\ifilter{r}{\istrf{s}}$ then $e:\istr$.
+  \item If $e=\istrf{s_1}+\istrf{s_2}$ then $e:\istr$.
+\end{itemize}
+\end{defn}
+
+For simplicity, we assume a fixed translation from
+$\lcs$ regular expressions to regular expressions recognizable by the internal
+language's regex library (in practice, a fixed translation is acceptable.)
+To summarize, we assume an internal language containing 
+a string type together with operations for string concatentation and filtering. We
+expect closure over strings for both operations.
+Finally, recall that ${\tt ifilter}$ is only needed for dynamic casts, which may be removed without
+descreasing the expressivity or even usability of the language.
+Finally, the semantics of the filter function are defined in terms of ${\tt rl\_filter}$,
+which is a static version of ${\tt ifilter}$.
+
+\subsection{Properties of Regular Languages}
+
+The regular languages are the smallest set generated by regular expressions
+defined in Figure 1.
+
+\begin{thm}{Closure Properties.} \label{thm:closure}
+The regular expressions are closed under complements and concatenation.
+\end{thm}
+\begin{proof}
+See \cite{cinderella}.
+\end{proof}
+
+\begin{thm}{Coercion Theorem.} \label{thm:coerce}
+Suppose that $R$ and $L$ are regular expressions, and that $s \in R$ is a finite string.
+Let $s' := coerce(R,L,s)$ with all maximal substrings recognized by $L$ replaced with $\epsilon$.
+Then $s'$ is recognized by $(R \backslash L) + \emptyset$ and the construction of $R \backslash L$ is decidable.
+\end{thm}
+\begin{proof}
+Let $F,G$ be FAs corresponding to $R$ and $L$, and let $G'$ be $G$ with its final states inverted (so that $G'$ is the complement of $L$).
+Define an FA $H$ as a DFA corresponding to the NFA found by combining $F$ and $G'$ such that $H$ accepts only if $R$ and $L'$ accept or if $s$ is empty (this construction may result in an exponential blowup in state size.)  Clearly, $H$ corresponds to $R \backslash L + \emptyset$.
+Thus, the construction of $R \backslash L + \emptyset$ is decidable.
+
+If $R \subset L$, $s' = \emptyset$.
+If $L \subset R$, either $s' = \emptyset$, or $s' \in R$ and $s' \not \in L$.
+If $R$ and $L$ are not subsets of one another, then it may be the case that $L$ recognizes part of $R$.
+Consider $L$ as the union of two languages, one which is a subset of $R$ and one which is disjoint.
+The subset language is considered above and the disjoint language is inconsequential.
+\end{proof}
+
+\onecolumn
+\begin{figure}
+\begin{center}
+\begin{tabular}{c r}
+%\inferline{T-VAR}{\Gamma \vdash x : T}{x:T \in \Gamma}
+%
+%\inferline{T-ABS}{\Gamma \vdash ({\tt \textbf{lambda}\ } x: t_2) : T_1 \rightarrow T_2}
+%			{\Gamma, x:T_1 \vdash t_2:T_2}
+%
+%\inferline{T-APP}
+%			{\Gamma \vdash t_1t_2 : T_{12}}
+%			{\Gamma \vdash t_1 : T_{11}\rightarrow T_{12} \\ \Gamma \vdash t_2 : T_{11}}	
+%
+\inferline{T-Equiv-Top}
+{\str \equiv \strin[.*]}
+{ }
+
+\inferline{T-Equiv-string\_in}
+{\strin[r] \equiv \strin[r']}
+{\lang{r} = \lang{r'}}
+
+\inferline{T-Equiv-Include}
+{\Gamma \vdash e:\strin[r]}
+{e \in \lang{r}}
+
+\inferline{T-Minimum}
+{ \Gamma \vdash e : \strin[e] }
+{\Gamma \vdash e : \str}
+
+%\inferline{T-Union}
+%  {\Gamma \vdash \stru[r'](e) : \strin[r \cup r']}
+%  { \Gamma \vdash e : \strin[r] }
+
+\inferline{T-Concat}
+{\Gamma \vdash e_1 + e_2 : \strin[r_1 \cdot r_2]}
+{ \Gamma \vdash e_1 : \strin[r_1] \\ \Gamma \vdash e_2 : \strin[r_2] }
+
+\inferline{T-Convert}
+{\Gamma \vdash [\strin[r']](e):\strin[r']}
+{\Gamma \vdash e : \strin[r] \\ \lang{r'} \subseteq \lang{r}}
+
+\inferline{T-Dconvert (optional)}
+{\Gamma \vdash \dconvert{\strin[r']}{e}:\strin[r']}
+{\Gamma \vdash e : \strin[r]}
+
+\inferline{T-Filter}
+{\Gamma \vdash \filter{\strin[r']}{t} : \strin[(r \backslash r') + \emptyset]}
+{\Gamma \vdash t:\strin[r]}
+
+\end{tabular}
+\caption{Typing relation for $\lcs$}
+\label{fig:lcsTyping}
+\end{center}
+\end{figure}
+
+%
+% Begin Operational Semantics
+%
+\begin{figure}
+\begin{center}
+\begin{tabular}{c r c r}
+%\inferline{E-App1}{t_1t_2 \Rightarrow t_1't_2}{t_1 \Rightarrow t_2}
+%\inferline{E-App2}{v_1t_2 \Rightarrow v_1t_2'}{t_2 \Rightarrow t_2'}
+%\inferline{E-AppAbs}{({\tt lambda } x : t_{12})v_2 \Rightarrow [x / v_2]t_{12}}{T_{11}}
+
+\inferline{E-strinval}
+  {\strf{s}:\strin[r] \val}
+  { }
+
+%\inferline{E-strval}
+%{\strf{s}:\str \val}
+%  { }
+
+\inferline{E-ival}
+  {\istrf{e}:\istr \ival }
+  { }
+
+\inferline{E-ifilterval}
+{ {\tt ifilter}(r,e) \ival }
+{ e {\tt ival} }
+
+\inferline{E-string}
+{e : \str \ireduces \istrf{e}}
+{ e \val }
+
+\inferline{E-string\_in}
+{e : \strin[r] \ireduces  \istrf{e}}
+{ e \val }
+
+\inferline{E-concatval}
+{e_1 + e_2 : \strin[r_1+r_2] \ireduces \istrf{e_1} + \istrf{e_2}}
+{e_1 \val \\ e_2 \val }
+
+\inferline{E-concatL}
+{e_1 + e_2 : \strin[r_1+r_2] \reduces {e_1}' + {e_2}}
+{ e_1 : \strin[r_1] \reduces {e_1}'}
+
+\inferline{E-concatR}
+{e_1 + e_2 : \strin[r_1+r_2] \reduces {e_1} + {e_2}'}
+{ e_2 : \strin[r_2] \reduces {e_2}'}
+
+\inferline{E-filterval}
+{\filter{\strin[r']}{e} : \strin[r\backslash r' + \emptyset] \reduces {\tt rl\_filter}(r,e)}
+{ e \val }
+
+\inferline{E-filter}
+{\filter{\strin[r']}{e} : \strin[r\backslash r' + \emptyset] \reduces  \filter{\strin[r]}{ e' }}     
+{ e : \strin[r] \reduces e' }
+
+\inferline{E-convertval}
+{ [\strin[r']](e) : \strin[r'] \ireduces \istrf{e} }
+{ e \val }
+
+\inferline{E-convert}
+{ [\strin[r']](e) : \strin[r'] \reduces [\strin[r']](e') }
+{ e : \strin[r] \reduces e' }
+
+\inferline{E-dconvertval (optional) }
+{\dconvert{\strin[r']}{e} : \strin[r'] \reduces {\dconvert{\strin[r']}{e'}}}
+{e:\strin[r] \reduces e'}
+
+\inferline{E-dconvert (optional) }
+{\dconvert{\strin[r']}{e} : \strin[r'] \ireduces \ifilter{r'}{\istrf{e}}}
+{e \val }
+
+\end{tabular}
+\caption{SOS rules for $\lcs$}
+\label{fig:lcsEval}
+\end{center}
+\end{figure}
+
+\twocolumn
+
+\subsection{Type Safety Proof}
+
+%\begin{lem}[Canonical Forms for $\lcs$]
+%We identify only canonical forms necessary for the type safety proof.
+%\begin{itemize}
+%  \item If $e \val$ and $e:\str$ then $e = \strf{s}$.
+%\end{itemize}
+%\end{lem}
+
+\begin{thm}[Preservation]
+  Let $T$ be a type in $\lcs$ and $h(T)=\sigma$ the corresponding type in the internal language.
+  For all terms $e$:
+  \begin{itemize}
+    \item If $e:T$ and $e:T \ireduces i$ then $i : \sigma$ such that $h(T) = \sigma$.
+    \item If $e:T$ and $e:T \reduces e'$ then $e':T$.
+  \end{itemize}
+\end{thm}
+\begin{proof}
+The proof is a straightforward induction on the derivation of the combined evaluation relation.
+\begin{itemize}[label=$ $,itemsep=1ex]
+  \item \textbf{E-Ival, E-Ifilterval}. Both cases hold since the terms at hand are not $\lcs$ terms.
+  \item \textbf{E-Stringval, E-strval}. Both cases hold since no reduction is possible. 
+  \item \textbf{E-String}. By the definition of typing for internal terms, $\istrf{e} : \istr$. It suffices to show that $h(\str) = \istr$, which follows from the definition of $h$.
+  \item \textbf{E-String\_in}.
+By the definition of typing for internal terms, $\istrf{e} : \istr$.
+It suffices to show that $h(\strin[r]) = \istr$, which follows from the definition of $h$ for arbitrary $r$.
+  \item \textbf{E-concatval}.
+By the definition of typing for internal terms, $\istrf{e_1} + \istrf{e_2} : \istr$.
+So it suffices to show that $h(\strin[r_1+r_2]) = \istr$, which follows from the definition of $h$ for arbitrary $r_1$, $r_2$.
+  \item \textbf{E-concatR, E-concatL}.
+Consider E-concatR.
+By induction, $e_1' : \strin{l_1}$.
+By inversion of T-Concat at the premise, $e_2 : \strin{r_2}$.
+Therefore, $e_1 + e_2 : \strin[r_1 + r_2]$.
+The left rule is symmetric.
+  \item \textbf{E-Filterval}. We have that $\filter{\strin[r']}{e}:\strin[r\backslash r' + \emptyset]$. By inversion of T-Filter, $e:\strin[r]$.
+    By T-Equiv-string\_in (which is bidirectional), $e \in \lang{r}$.
+    By the Coercion Theorem, ${\tt rl\_filter}(r,e) \in \lang{r\backslash r' + \emptyset}$.
+    By T-Equiv-string\_in,  $e \in \lang{r}$ and ${\tt rl\_filter}(r,e) \in \lang{r\backslash r' + \emptyset}$ implies ${\tt rl\_filter}(r,e):\strin[r\backslash r' + \emptyset]$.
+  \item \textbf{E-Filter}. By inversion, $e : \strin[r']$ $\reduces e'$ so by the induction $e':\strin[r']$.
+    Therefore, \\${\tt filter}(\strin[r], e'):$ $\strin[r \backslash r' + \emptyset]$ by T-Filter.
+  \item \textbf{E-Convertval}. It suffices to show that $h(\strin[r]) = \istr$, which is true by definition.
+  \item \textbf{E-Convert}. By inversion and induction, $e' : \strin[r]$. We know that $[\strin[r']](e) : \strin[r']$, so by inversion of T-Convert $\lang{r'} \subseteq \lang{r}$.
+    It follows that $[\strin[r']](e') : \strin[r']$.
+  \item \textbf{E-DConvert}. By inversion, $e : \strin[r'] \reduces e'$. By the induction hypothesis, $e' : \strin[r']$; therefore, by T-Dconvert, $\dconvert[\strin[r]](e):\strin[r]$.
+  \item \textbf{E-DConvertval}.
+By the definition of typing for internal terms, $\ifilter(r,\istrf{e}) : \istr$.
+It suffices to show that $h(\strin[r_1+r_2]) = \istr$, which follows from the definition of $h$.
+\end{itemize}
+\end{proof}
+
+\begin{thm}[Progress]
+  If $e:T$ then $e:T \reduces^* e' \ireduces^* i$ where $i \ival$.
+\end{thm}
+\begin{proof}
+By induction on the derivation of $e:T$.
+For \textbf{T-Equiv-Include}, note that $e \in \lang{r}$, $e = \strf{s}$ for some $s$; therefore, $e \val$ by E-Strinval. 
+We have that $e \val$ and $e : \strin[r]$, so $e \ireduces \istrf{e}$ by E-string\_in.
+The remaining cases follow by induction on the hypotheses and application of corresponding evaluation rules.
+\end{proof}
+
 \section{Implementation Inside a Python}
+
+TODO-nrf rewrite
+The $\lcs$ language is implemented as a type system extension in Ace; this extension
+is illustrated in the examples at the beginning of this paper.
+
+Computing a regular expression representating the language
+$R \backslash S$ is necessary in order to type-check expressions in which the filter function
+occurs. This language is computed by translating $R$ and $S$ into finite automata,
+complementing the final states of $S$, then constructing the cross-product of $R$
+and $S'$. Type checking terminates only because this construction is decidable.
+
+In addition to this construction, other more typical operations -- such
+as equality checks for regular expressions and regular expression matching -- are
+also necessary. For these reasons, the Ace extension implementing $\lcs$ includes a library
+implementing each of these constructions.
+
 \subsection{Background: Ace}
@@
 \end{document}
+
+\end{document}

### 2014-04-13 sanitation-plas14/sanitation-plas14.tex

@@
 In web applications and other settings, incorrect input sanitation often causes
-security vulnerabilities. For this reason, frameworks often provide methods
-for sanitizing user input. When these methods are insufficient or unavailable,
-developers often create  custom input sanitation algorithms. In both cases,
-input sanitation techniques are ultimately implemented using the language's
-regex library.
+security vulnerabilities. In fact, ...OWASP says it's important... .
+Mention CVE stats.
+For this reason, modern web frameworks and libraries use various techniques to
+ensure proper sanitation of arbitrary user input. On the rare occasions when
+these methods are unavailable or insufficient, developers (hopefully) create 
+ad hoc sanitation algorithms. Im most cases, sanitation algorithms -- ad hoc 
+or otherwise -- are ultimately implemented using the language's regular
+expression capabilities. Therefore, a system capable of statically 
+checking properties about operations performed using regular expressions will
+be expressive enough to capture real-world implementations of sanitation 
+algorithms.
 
-Formally, input sanitation is the problem of ensuring that an arbitrary string
-is converted into a safe form before potentially unsafe use. For example, consider
-SQL injection attacks. To prevent such attacks, we might ensure that any string
-used as input to a query does not contain unescaped SQL command sequences.
+Input sanitation is the problem of ensuring that an arbitrary string
+is coerced into a safe form before potentially unsafe use. For example, 
+preventing SQL injection attacks requires ensuring that any string
+coming from user input to a query does not contain unescaped SQL. 
+The point at which arbitrary user input is concatenated into a SQL query is called a use site.
+Although we believe are general approach extends to a wider class of problems
+(e.g. sanitation algorithms for preventing XSS attacks might be definable using
+regular tree languages), these generalizations are beyond the scope of the 
+present paper.
 
-This appendix presents a type system, $\lcs$, for ensuring that input 
-sanitation algorithms are implemented correctly with respect to use site 
-specifications.
+This paper presents a language extension, definable in the Ace programming language,
+for ensuring that input sanitation algorithms are implemented correctly with
+respect to use site specifications. If use site specifications are sufficient, 
+then type checking ensures the absence of vulnerabilities and other bugs which
+arise from improper sanitation. Our extension focuses on sanitation algorithms 
+which aim to prevent command-injection style attacks (e.g. SQL injection or RFI).
 
-Unlike frameworks provided by general purpose languages such as Haskell and
+\subsection{Related Work and Alternative Approaches}
+
+The input sanitation problem is well-understood. There exist a large number
+of techniques and technologies, proposed by both practitioners and researchers, for
+preventing injection-style attacks. In this section, we defend the novelty and
+significance of our approach with respect to the state of the art in practice 
+and in research.
+
+Unlike frameworks provided by languages such as Haskell and
 Ruby, our type system provides a \emph{static} guarantee that input is always 
 properly sanitized before use. We achieve this by defining a typing relation
-which captures idiomatic sanitation algorithms. Type safety relies upon 
+which captures idiomatic sanitation algorithms. Type safety in our system relies upon 
 several closure and decidability results about regular languages.
 
-XDuce typechecks XML. Like our work, XDuce relies upon some properties of
-regular expressions to establish soundness and completeness results. Unlike
-XDuce, our work is motivated by input sanitation and therefore considers arbitrary
-strings (as opposed to tree-structured XML documents). Furthermore, our static
-treatment of the ubiquitous regex filter (replace) function for regular expression is novel.
+Libraries and frameworks available in functional programming language communities often
+make claims about security and sometimes even sophistically mention sophisticated
+type systems as evidence of freedom from injection-style attacked.
+However, even the specification that input is always sanitized properly before use is not
+actually captured by the type system or anywhere else; in fact, the full specification of these
+algorithms is rarely characterized by anything more specific than the type
+String $\rightarrow$ String, which is a class of safe input sanitation algorithms
+only in the most degenerate case.
 
-Finally, some research languages achieve similar safety guarantees via a specialized type
-system. The existence of such domain-specific type systems is suggestive, and 
-we believe the simplicity of $\lcs$ demonstrates the effectiveness of
-extensibility. Instead of introducing entirely new languages for each domain,
-language extension developers may invest incrementally in obtaining static
-guarnatees which address common mistakes.
+A number of research languages provide actually static guarantees that a program
+is free of input sanitation vulnerabilities. Most rely on some form of information
+flow. TODO-nrf  give citations and examples. Our extension to Ace differs from
+these systems in the ways following:
+\begin{itemize}
+  \item Our system is a light-weight solution to a single class of sanitation
+  vulnerabilities (e.g. we do not address Cross-Site Scripting). We present our
+  system not as a comprehensive solution to the web security problem, but rather as
+  evidence that composable, light-weight and simple analyses can address security
+  problems.
+  \item Our system is defined as a library in terms of an extensible type system,
+  as opposed to a stand-alone language. This is important for two reasons.
+  First, our system is one of the first large examples written in Ace, and serves
+  as an extended case study. Second, although our approach requires
+  developers to adopt a new language, it does not require developers to adopt
+  a language specifically for web development.
+  \item Ace is implemented in Python and shares its grammar. Since Python is a 
+  popular programming language among web developers, the barrier between our research
+  and adopted technologies is far lower than is e.g. Ur/Web's.
+  \item In general, our is a composable, light-weight and
+  natural solution to a single problem -- rather than a comprehensive 
+  solution to the web security problem. Our goal is to demonstrate that extensible
+  type systems can capture static properties of common idioms, and thereby ensure
+  safety without introducing much additional complexity.
+\end{itemize}
+
+Finally, incorporating regular expressions into the type system is not novel.
+The XDuce system \cite{pierce} typechecks XML schemas using regular expressions.
+We differ from this and related work in at least two ways.
+First, our system  is defined within an extensible type system;
+this first difference introduced an interesting class of design problems (see \S ???).
+TODO-nrf this is the second time this is used as a warrant. Factor out this argument
+and place it at the top???
+Second, our focus on security required novel design decisions; for instance, the
+filter elimination form is unique to Ace.
+
+In conclusion, our system is novel in at least two ways:
+\begin{itemize}
+\item The safety guarantees provided by libraries and frameworks in popular languages
+are not as (statically) justified as is often belived (or even claimed).
+\item Our extension is the first major demonstration of how an extensible type
+system may be used to provide light-weight, composable security analyses based upon
+idiomatic code.
+\end{itemize}
+
+\subsection{Outline}
 

### 2014-04-14 thesis-proposal/problem.tex
Final draft of thesis proposal

@@
 %%One might exp ``generality''
-Specifying and implementing a simple but general-purpose programming language and its associated tools (collectively, a \emph{programming system}) has long been seen as a grand challenge in computing.
-Were an ideal system to emerge, we might expect that providing a new high-level abstraction would not require specifying a new dialect of the language or its tools, because a library-based embedding of the abstraction would be essentially as reasonable, natural and performant.
-By this metric, we are clearly far from our ideal.
-New language and tool dialects, originating both in research and in practice and spanning all major language lineages, continue to be widely used as vehicles for new abstractions (examples of which we will discuss throughout this work).
-This suggests that not all useful  abstractions can be seen, when considered comprehensively, as mere modes of use of a contemporary ``general-purpose'' programming system.% The problem of achieving both simplicity and generality simultaneously with a single system remains unsolved.
+Specifying and implementing a ``general purpose'' programming language and its associated tools (collectively, a \emph{programming system}) has long been a grand challenge in computing.
+Were a truly general system to emerge, we might expect that its users would rarely  need to develop a new dialect of the system.
+Embeddings of new abstractions using mechanisms built into the language would be suitably reasonable, natural and performant.
+Alas, new language and tool dialects spanning all major language lineages continue to be widely used as vehicles for new abstractions (examples of which we will discuss throughout this work), suggesting that not all useful  abstractions can be seen, when considered comprehensively, as mere modes of use of any contemporary ``general purpose'' programming system.% The problem of achieving both simplicity and generality simultaneously with a single system remains unsolved.
 
-We observe that new dialects are often motivated by the desire to introduce new syntax, strengthen the language's type system or provide specialized editor services, all aspects of existing systems that appear rather monolithic when considered from the perspective of a library provider.
+We observe that new dialects are often motivated by the desire to introduce new syntax, strengthen the language's static semantics or provide specialized editor services, all aspects of existing systems that appear fixed when considered from within (that is, from the perspective of a library provider).
 %, encouraged  historically  by the availability of tools like compiler generators and,  more recently, language workbenches \cite{workbenches} and DSL frameworks \cite{dsl}.
 Unfortunately, taking this approach makes it substantially more difficult for clients to import high-level abstractions orthogonally.
-It follows, then, that {language-integrated mechanisms} that allow library providers to extend these features of the system would increase its generality by decreasing the need for new language and tool dialects.
-However, such mechanisms must be introduced with care, because decentralizing control over such fundamental aspects of the system could destabilize it by weakening its metatheory and permitting ambiguities and conflicts (collectively, \emph{safety issues}).
-It could also make it more difficult to reason about the meaning of code.
-In this thesis, we aim to show that by adopting the natural convention of organizing language extensions around user-defined types (forming what we call \emph{active types}) and designing an underlying type system that enforces critical  abstraction barriers between extensions, we can give abstraction providers the ability to express from within libraries powerful, orthogonal extensions to the syntax and type system, and define powerful editor services, without the possibility of the kinds of safety issues that have limited prior approaches.
+It follows then that {language-integrated mechanisms} that allow library providers to extend these features of the system could strengthen its claim to generality by decreasing the need for new language and tool dialects.
+However, such mechanisms must be introduced with care, because decentralizing control over such fundamental aspects of the system could weaken its metatheory and permit ambiguities and conflicts when two libraries are combined (collectively, \emph{safety issues}).
+It could also make it more difficult to define new functions over expression forms in the language (the \emph{expression problem}) and make it more difficult to  understand the meaning of programs.
+In this thesis, we aim to show that by associating system extensions with types (forming what we call \emph{active types}) rather than new expression forms, and using a \emph{bidirectional type system} that enforces critical  abstraction barriers between extensions, we can  give abstraction providers the ability to orthogonally implement new syntax, type systems and editor services from within libraries, rather than as new language and tool dialects, while avoiding safety issues and retaining  an essentially conventional typing discipline.
 %We call user-defined types that introduce new features into the system in this way \emph{active types}.
@@
 DNA sequences) \cite{Thompson:1968:PTR:363347.363387}.
-If a programming system provided full support for this abstraction, it might  provide features like:
+Programmers who work with regular expressions regularly might want a programming system supporting features like:
 
@@
 \item \textbf{Syntax for pattern literals}.
-An ideal syntax would permit us to express patterns concisely and clearly.
+An ideal syntax would permit us to express patterns in a concise, conventional manner.
 For example, The BisI restriction enzyme cuts DNA whenever it sees the pattern \texttt{GC\textit{N}GC}, where \texttt{\textit{N}} represents any base.
@@
 %The cognitive load of reading and writing patterns is low and patterns are parsed once at compile-time. 
-Malformed patterns would result in intelligible {compile-time} errors.
+We would want malformed patterns to result in intelligible {compile-time} errors.
 \item A \textbf{type system} that ensures that key invariants related to regular expressions are statically maintained:
@@
 %By performing this step ahead-of-time, these dangers can be avoided.
-\item \textbf{Editor services} to support syntax highlighting of patterns and to allow clients to refer to documentation and interactively test patterns against examples strings (e.g.
-\cite{IntelliJRegexp}) or even extract patterns from examples interactively (e.g.
-as in \cite{_txt2re:_????}).
+\item \textbf{Editor services} to support syntax highlighting, documentation, interactive testing and pattern extraction from example strings.
 \end{enumerate}
@@
 %In a conventional \emph{monolithic} programming system, support for each of these features would need to be built into the language and tools. 
-No system today builds in support for all of the features enumerated above in their strongest form, so library providers must provide support for regular expressions by leveraging  general-purpose constructs.
-Unfortunately, it is impossible to completely define the syntax and the specialized static semantics described in the references above in terms of general-purpose notations and abstractions, so library providers need to compromise.
+No system today builds in support for all of the features enumerated above in their strongest form, so library providers must provide support for regular expressions by leveraging  general-purpose mechanisms.
+Unfortunately, it is impossible to completely define the syntax and the specialized static semantics referenced above in terms of general-purpose notations and abstractions, so library providers need to compromise.
 
 The most common strategy is to ask clients to enter patterns as strings, deferring their parsing, typechecking, compilation and use all to run-time.
-This provides only a weak approximation to feature 1 (due to clashes between string escape sequences and regular expression syntax).
-None of the static guarantees can be provided in this way, leading to run-time exceptions (even in well-tested code \cite{spishak2012type}), and logic errors that cannot easily be caught even at run-time, some of which can lead to security vulnerabilities (due to injection attacks \cite{owasp2013}).
-It also introduces performance overhead due to run-time parsing and compilation of patterns, and, in the absence of feature 2c,   redundant run-time string checks.
-Requiring that clients instead introduce patterns directly using general-purpose constructs like object or datatype constructors or operations over an abstract type rather than via strings provides only feature 2a at the expense of the weak approximation to feature 1 that the use of string literals provided.
+This provides only a partial approximation to feature 1 (due to clashes between string escape sequences and regular expression syntax).
+None of the static guarantees can be provided in this way, leading to run-time exceptions (even in well-tested code \cite{spishak2012type}), and logic errors that cannot even be caught at run-time, some of which can lead to security vulnerabilities (due to injection attacks \cite{owasp2013}).
+It also introduces performance overhead due to run-time parsing and compilation of patterns, and redundant run-time checks.
+
+Requiring that clients instead introduce patterns directly using general-purpose constructs like sums or products (as exposed by, e.g., functional datatypes or objects) or operations over an abstract type, rather than via strings, provides only feature 2a at the expense of the weak approximation to feature 1 that the use of string literals provided.
 
@@
 Regular expressions are not trivial to work with, and we have found that tool support can be quite helpful, even for professional programmers \cite{Omar:2012:ACC:2337223.2337324}.
-A number of tools are available online that provide helpful services like syntax highlighting for patterns and pattern extraction from examples (e.g.
+Indeed, a number of tools are available online that provide services like syntax highlighting, testing and explanations of patterns \cite{regexr} and pattern extraction from examples (e.g.
 \cite{_txt2re:_????}).
 Unfortunately, tools that must be accessed externally are difficult to discover and are thus used infrequently \cite{Murphy-Hill:2011:PIE:1958824.1958888,Campbell:2008:DRT:1636642.1636651,Omar:2012:ACC:2337223.2337324}.
-We found in experiments that they are also less usable than editor-integrated tools %(e.g.
+We found that they are also less usable than editor-integrated tools %(e.g.
 \cite{IntelliJRegexp})
 because they require the programmer to switch contexts  and cannot make use of the code context \cite{Omar:2012:ACC:2337223.2337324}.
-Working with high-level abstractions in a simple editor is thus awkward and error-prone, just like attempting to work with high-level abstractions in a low-level language.
-For these reasons,   we consider editor behavior as within the purview of an abstraction specification.
+Working with high-level abstractions in a ``low-level'' editor is thus awkward and error-prone, just like attempting to work with high-level abstractions in a low-level language.
+For these reasons,   we consider editor behavior as within the scope of an abstraction's specification.
 \subsection{Language-External Approaches}\label{external-approaches}
@@
 \caption{\small (a) When taking a language-external approach, new features are packaged together into separate languages and tools, causing problems with orthogonality and client compatibility (described in the text).
-(b) When taking a language-integrated approach, there is one extensible host language and the compile-time and edit-time logic governing new constructs is expressed within ``active'' libraries.
-If the extension mechanism guarantees that extensions cannot cause instabilities, the problems of orthogonality and client compatibility are avoided.}
+(b) When taking a language-integrated approach, there is one extensible host language and the compile-time and edit-time logic governing new constructs is expressed within ``active libraries''.
+If the extension mechanism precludes conflicts by construction, the problems of orthogonality and client compatibility are avoided.}
 \label{approaches}
@@
 This results in a collection of distinct systems, as diagrammed in Figure \ref{approaches}a.
-This might be entirely sufficient if developing proofs of concept is the only goal, but when providers of new features take language-external approaches like this, it causes  problems for clients (including other researchers) related to \textbf{orthogonality} and \textbf{client compatibility}.
-This has limited the broad adoption of these kinds of innovations.%This latter method couples the semantics of the feature to the implementation details of a particular tool.
+This might be entirely sufficient if developing a proof of concept is the only goal, but it makes it difficult to achieve adoption of these abstractions (even by other researchers) related to \textbf{orthogonality} and \textbf{client compatibility}.% This has limited the broad adoption of these kinds of innovations.%This latter method couples the semantics of the feature to the implementation details of a particular tool.
 Because the use of one implementation entails a different semantics for the feature than another, the extended tool acts, \emph{de facto}, as a distinct system for our purposes.
@@
 Recent evidence indicates that this is one of the major barriers preventing research from being driven into practice.
-For example, developers prefer language-integrated parallel programming abstractions with stronger safety guarantees and more natural syntax to library-based abstractions when all else is equal \cite{cave2010comparing}, but library-based implementations are more widely adopted because ``parallel programming languages'' privilege only a few chosen  abstractions at the language level.
-This is problematic because different parallel programming abstractions are seen as more appropriate in different situations \cite{Tasharofi:2013rc}.
-Moreover,  parallel programming support is rarely the only concern relevant to clients outside of a classroom setting.
+For example, developers prefer language-integrated parallel programming abstractions with stronger semantic guarantees, more efficient implementation and more natural syntax to library-based approximations when all else is equal \cite{cave2010comparing}, but library-based approximations are more widely adopted because  ``parallel programming languages'' privilege only a few chosen  abstractions at the language level.
+This is problematic, because different abstractions are seen as more appropriate in different situations \cite{Tasharofi:2013rc}.
+Moreover,  parallel programming is rarely the only concern relevant to clients outside of a classroom setting.
 Support for regular expressions, for example, would be simultaneously desirable for processing large amounts of genomic data in parallel, but using these features together in the same compilation unit would be difficult or impossible.
@@
 
-\paragraph{Client Compatibility} Even in cases where, for each component of a software system, there is a programming system considered entirely satisfactory by its developers, there remain problems at the interface between components.
+\paragraph{Client Compatibility} Even in cases where, for each component of a software system, there was a programming system considered entirely satisfactory by its developers, there would remain a problem at the interface between its components.
 An interface that exposes externally a specialized construct particular to one language (e.g.
@@
 
-One strategy often taken by proponents of a {language-oriented approach} to abstraction development \cite{journals/stp/Ward94} to partially address the client compatibility problem is to  target an established intermediate language and use its constructs as a common language for communication between components written in different languages.
+One strategy taken by proponents of the {language-oriented approach} to abstraction development \cite{journals/stp/Ward94} to partially address the client compatibility problem is to  target an established intermediate language and use its constructs as a common language for communication between components written in different languages.
 Scala \cite{200464/IC} and F\# \cite{pickering2007foundations} are examples of prominent general-purpose languages that have taken this approach, and most DSL frameworks also rely on this strategy.
@@
 %But many of the most innovative constructs found in modern languages (often, those that justify their creation) are difficult to define in terms of existing constructs in ways that guarantee all necessary invariants are statically maintained and that do not require large amounts boilerplate code and run-time overhead.
-As a simple example with significant contemporary implications, F\#'s type system does not admit \verb|null| as a value for a type introduced by F\#, but maintaining this sensible internal invariant still requires dynamic  checks because the stricter typing rules of F\# do not apply when F\# data structures are constructed by other languages on the Common Language Infrastructure (CLI) like C\# or SML.NET.
+As a simple example with significant contemporary implications, F\#'s type system does not admit \verb|null| as a value for any type both defined and used within F\# code, but maintaining this sensible internal invariant still requires dynamic  checks because the stricter typing rules of F\# do not apply when F\# data structures are constructed by other languages on the Common Language Infrastructure (CLI) like C\# or SML.NET.
 This is not an issue exclusive to intermediate languages that make regrettable choices regarding \verb|null|, however.
@@
 SML.NET prohibits exposing such types at component boundaries altogether.
-It cannot it naturally consume F\# data structures, despite having a rather similar syntax and semantics in many ways (both languages directly descend from ML).
+It cannot naturally consume F\# data structures, despite having a rather similar syntax and semantics in most ways (both languages directly descend from ML).
 %In Scala, traits that have default method implementations are difficult to implement from Java or other JVM languages and the workaround can break if the trait is modified \cite{scalatraitinterop}. 
@@
 More specifically, we will show how control over aspects of the \textbf{syntax}, \textbf{type system} and \textbf{editor services} can be delegated to user-defined logic distributed in {libraries}, as illustrated in Figure \ref{approaches}b.
-Such libraries have been called \emph{active libraries}  \cite{activelibraries} because, rather than being passive clients of features already available in the system, they contain logic invoked by the system during development or compilation to provide new features.
+Such libraries have been called \emph{active libraries}  \cite{activelibraries} because, rather than being passive clients of features already available in the system, they contain logic invoked by the components of the programming system while a program is being edited or compiled to provide new features.
 Features implemented within active libraries can be imported as needed, unlike features implemented by external means, seemingly avoiding the problems of orthogonality and client compatibility.
@@
 Resolving these kinds of ambiguities requires significantly more expertise with parser technology than using the syntax itself does.
-These and more serious safety and ambiguity issues have plagued previous attempts to design language-integrated extensibility mechanisms.
+These (and other more serious) safety and ambiguity issues have plagued previous attempts to design language-integrated extensibility mechanisms.
 We will briefly review some of these attempts below.% To prevent them, our mechanisms will organize  extension logic around types to guarantee that extensions are both safe in isolation and also safely composable in any combination.

### 2014-04-14 thesis-proposal/timeline.tex
Final draft of thesis proposal

--- 2014-01-16 thesis-proposal/timeline.tex
@@
 % !TEX root = omar-thesis-proposal.tex
+\section{Conclusion}\label{conclusion}
+In arguing for language extensibility, we are accepting the philosophy of logical pluralism, explained by Carnap as follows \cite{carnap:1950b}:
+\begin{quote}
+Let us grant to those who work in any special fields of investigation the freedom to use any form of expression which seems useful to them.
+The work in the field will sooner or later lead to the elimination of those forms which have no useful function.
+Let us be cautious in making assertions and critical in examining them, but tolerant in permitting linguistic forms.
+\end{quote}
+But we take issue with this unbounded \emph{principle of tolerance}, which he explained more directly as follows \cite{CarnapR:logsl}:
+\begin{quote}
+In logic, there are no morals.
+Everyone is at liberty to build his own logic, i.e.
+his own form of language, as he wishes.
+All that is required of him is that, if he wishes to discuss it, he must state his methods clearly, and give syntactic rules instead of philosophical arguments.
+\end{quote}
+
+Carnap formulated his principle of tolerance because he wished to freely explore the consequences of reasoning with different connectives and  rules and emphasize that conducting philosophical reasoning by methodical logical analysis did not require one to accept any particular axioms.
+But he was also a prominent member of the Vienna circle, which sought a ``unified science'' the purpose of which was "to link and harmonise the achievements of individual investigators in their various fields of science" by the construction of a "constitutive system" \cite{hahn1929scientific}.
+These two programs were in conflict: were the Vienna circle's ``constitutive system'' to become na\"ively tolerant of the inference rules (what Carnap called L-rules) developed by individual investigators, it would find that this would then deny them autonomy of reasoning.
+According to social scientist Raz, the primary justification of tolerance is to maximize
+individual autonomy in a multi-party context \cite{raz1988autonomy}.
+Yossie Nehushtan argues that a liberal  society ``should not tolerate anything that denies the justifications of tolerance'' \cite{nehushtan2007limits}.
+Thus, we propose a more cooperative \emph{principle of reciprocal tolerance}:
+\begin{quote}
+In metalogic, there \emph{are} morals.
+Everyone is at liberty to build their own logical fragment, i.e.
+their own language extensions, as long as these can be implemented within a framework that ensures, by formal means instead of philosophical argument, that autonomously derived reasoning principles are conserved when fragments are combined, in \emph{any} combination.
+\end{quote}
+
+By following this principle, we believe that a research program that supports Carnap's logical pluralism within a ``constitutive system'' of the sort envisioned by the Vienna circle is both possible and practical.
+The work in this thesis gives first steps towards this goal.
+
+%Aside from the archaic use of gendered pronouns, 
+% If the reader will permit a brief digression into social science, the adoption of a tolerant viewpoint is difficult to sustain in a society that does not require \emph{reciprocity}.
+And indeed, this was precisely Carnap's program.
+But
+
+% http://www.trinitinture.com/documents/nehushtan.pdf
+% https://en.wikipedia.org/wiki/Vienna_Circle
+% http://consequently.org/papers/carnap.pdf
+% https://www.academia.edu/4041290/How_tolerant_can_you_be_Carnap_on_the_normativity_of_logic
+
 \section{Timeline}
+To summarize, we plan on completing the work in this thesis per the following schedule:
 \begin{itemize}
-\item @$\lambda$ - ICFP - Mar. 1
-\item Wyvern - ECOOP or OOPSLA - late March
-\item Ace - PLDI or OOPSLA - late March
-\item Then write everything up.
+\item The work on type-specific languages has been accepted to ECOOP 2014.
+The deadline for final revisions is May 12, so we will complete the remaining work, including writing a technical report, by then.
+This should be sufficient for the chapter on Wyvern in the dissertation.
+\item The work on @$\lambda$ and active embeddings is conceptually complete.
+We plan to complete the technical details and exposition and submit it to POPL in July.
+This should be sufficient for the chapter on @$\lambda$ in the dissertation.
+\item The work described the core of Ace was submitted to OOPSLA 2014 on Mar.
+25.
+Notification is May 26.
+A paper with Nathan Fulton, a former undergraduate student who I co-supervised, on regular expression types in Ace is in preparation for submission to PLAS 2014 on April 20.
+We plan to submit a paper to either SLE 2014 or GPCE 2014 on FFIs using Ace (in particular, a FFI for OpenCL, which was already developed); both conferences have due dates at the end of May.
+Together, these will form the chapters on Ace in the dissertation.
+\item The work on active code completion has been completed and published at ICSE 2012 \cite{Omar:2012:ACC:2337223.2337324}.
+The remaining work is expository and will be completed in the course of writing the Graphite chapter of the dissertation.
 \end{itemize}

### 2014-04-14 sanitation-plas14/sanitation-plas14.tex

--- 2014-04-09 sanitation-plas14/sanitation-plas14.tex
@@
+%TODO-nrf:
+% * Type system definition -- check big step semantics
+% PRoofs tonigt:
+% * Type safety proof for internal language
+% * Representational consistency
+
 %\documentclass[12pt]{article}
@@
 
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\usepackage{bussproofs}    % Gentzen-style deduction trees *with aligned sequents!*
+\usepackage{mathpartir}	% For type-settting type checking rule figures
+\usepackage{syntax}		% For type-setting formal grammars.
+
+\usepackage{hyperref}		% For links in citations
+
+\usepackage{float}			% Make figures float if [H] option is passed.
+
+\iffalse
+\usepackage{listings}		% For typesetting code listings
+\usepackage{callout}
+\usepackage{titlesec}
+\usepackage[T1]{fontenc}
+\usepackage{upquote}
+\lstset{upquote=true}
+\fi
+
+\usepackage{textcomp}		% For \textquotesingle as used in introduction
+\usepackage{color}			% for box colors, like in TAPL.
+
+\usepackage{amsmath}		% Begin Carthage default packages
+\usepackage{makeidx}
+\usepackage{amsthm}
+\usepackage{amsfonts}
+\usepackage{amssymb}
+\usepackage{graphics}
+\usepackage{enumerate}
+\usepackage{multicol}
+\usepackage{epsfig}
+\usepackage{csquotes}
+\usepackage{enumitem}
+
+\newtheorem{thm}{Theorem}                                                       
+\newtheorem{cor}[thm]{Corollary}                                                
+\newtheorem{lem}[thm]{Lemma}                                                    
+\newtheorem{prop}[thm]{Proposition}                                             
+\newtheorem{ax}[thm]{Axiom}                                                     
+\theoremstyle{definition}                                                       
+\newtheorem{defn}[thm]{Definition}                                              
+\newtheorem{exam}[thm]{Example}                                                 
+\newtheorem{rem}[thm]{Remark}
+
+%
+% For type setting inference rules with labels.
+%
+\newcommand{\inferlbl}[3]
+			{\inferrule{#3}{#2}{\textsf{\footnotesize{\sc #1}}}}
+\newcommand{\inferline}[3]
+			{\inferrule{#3}{#2} & {\textsf{\footnotesize{\sc #1}}} \\ \\}
+
+\newcommand{\Lagr}{\mathcal{L}}
+\newcommand{\strin}{{\tt string\_in}}
+\newcommand{\lang}[1]{\Lagr\{#1\}}
+\newcommand{\stru}[2]{ {\tt string\_union}(#1,#2)}
+
+\newcommand{\str}{ {\tt string} }
+
+\newcommand{\dconvert}[2]{ {\tt dconvert}(#1,#2) }
+\newcommand{\filter}[2]{ {\tt filter}(#1,#2) }
+\newcommand{\ifilter}[2]{ {\tt ifilter}(#1,#2) }
+
+\newcommand{\reduces}{ \Rightarrow }
+\newcommand{\ireduces}{ \Downarrow }
+\newcommand{\sreduces}{ \leadsto }
+\newcommand{\gvd}{\Gamma \vdash }
+\newcommand{\ovd}{\Omega \vdash }
+
+\newcommand{\trep}{{\tt rep}}
+
+\newcommand{\val}{ \ {\tt val} }
+\newcommand{\ival}{ \ {\tt ival} }
+
+\newcommand{\istrf}[1]{`#1\textrm'} %??
+\newcommand{\strf}[1]{``#1"}
+
+%
+% Constrain the size of full-page diagrams and rule lists
+%
+%\newcommand{\pagewidth}{5in}
+%\newcommand{\rulelistwidth}{3in}
+
+%
+% Names of type systems presented in paper
+%
+\newcommand{\lcs}{\lambda_{CS}}
+
+\setlength{\grammarindent}{3em}
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
 \begin{document}
@@
 \section{Introduction}\label{intro}
-INTRO HERE
+In web applications and other settings, incorrect input sanitation often causes
+security vulnerabilities. In fact, ...OWASP says it's important... .
+Mention CVE stats.
+For this reason, modern web frameworks and libraries use various techniques to
+ensure proper sanitation of arbitrary user input. On the rare occasions when
+these methods are unavailable or insufficient, developers (hopefully) create 
+ad hoc sanitation algorithms. Im most cases, sanitation algorithms -- ad hoc 
+or otherwise -- are ultimately implemented using the language's regular
+expression capabilities. Therefore, a system capable of statically 
+checking properties about operations performed using regular expressions will
+be expressive enough to capture real-world implementations of sanitation 
+algorithms.
+
+Input sanitation is the problem of ensuring that an arbitrary string
+is coerced into a safe form before potentially unsafe use. For example, 
+preventing SQL injection attacks requires ensuring that any string
+coming from user input to a query does not contain unescaped SQL. 
+The point at which arbitrary user input is concatenated into a SQL query is called a use site.
+Although we believe are general approach extends to a wider class of problems
+(e.g. sanitation algorithms for preventing XSS attacks might be definable using
+regular tree languages), these generalizations are beyond the scope of the 
+present paper.
+
+This paper presents a language extension, definable in the Ace programming language,
+for ensuring that input sanitation algorithms are implemented correctly with
+respect to use site specifications. If use site specifications are sufficient, 
+then type checking ensures the absence of vulnerabilities and other bugs which
+arise from improper sanitation. Our extension focuses on sanitation algorithms 
+which aim to prevent command-injection style attacks (e.g. SQL injection or RFI).
+
+\subsection{Related Work and Alternative Approaches}
+
+The input sanitation problem is well-understood. There exist a large number
+of techniques and technologies, proposed by both practitioners and researchers, for
+preventing injection-style attacks. In this section, we defend the novelty and
+significance of our approach with respect to the state of the art in practice 
+and in research.
+
+Unlike frameworks provided by languages such as Haskell and
+Ruby, our type system provides a \emph{static} guarantee that input is always 
+properly sanitized before use. We achieve this by defining a typing relation
+which captures idiomatic sanitation algorithms. Type safety in our system relies upon 
+several closure and decidability results about regular languages.
+
+Libraries and frameworks available in functional programming language communities often
+make claims about security and sometimes even sophistically mention sophisticated
+type systems as evidence of freedom from injection-style attacked.
+However, even the specification that input is always sanitized properly before use is not
+actually captured by the type system or anywhere else; in fact, the full specification of these
+algorithms is rarely characterized by anything more specific than the type
+String $\rightarrow$ String, which is a class of safe input sanitation algorithms
+only in the most degenerate case.
+
+A number of research languages provide actually static guarantees that a program
+is free of input sanitation vulnerabilities. Most rely on some form of information
+flow. TODO-nrf  give citations and examples. Our extension to Ace differs from
+these systems in the ways following:
+\begin{itemize}
+  \item Our system is a light-weight solution to a single class of sanitation
+  vulnerabilities (e.g. we do not address Cross-Site Scripting). We present our
+  system not as a comprehensive solution to the web security problem, but rather as
+  evidence that composable, light-weight and simple analyses can address security
+  problems.
+  \item Our system is defined as a library in terms of an extensible type system,
+  as opposed to a stand-alone language. This is important for two reasons.
+  First, our system is one of the first large examples written in Ace, and serves
+  as an extended case study. Second, although our approach requires
+  developers to adopt a new language, it does not require developers to adopt
+  a language specifically for web development.
+  \item Ace is implemented in Python and shares its grammar. Since Python is a 
+  popular programming language among web developers, the barrier between our research
+  and adopted technologies is far lower than is e.g. Ur/Web's.
+  \item In general, our is a composable, light-weight and
+  natural solution to a single problem -- rather than a comprehensive 
+  solution to the web security problem. Our goal is to demonstrate that extensible
+  type systems can capture static properties of common idioms, and thereby ensure
+  safety without introducing much additional complexity.
+\end{itemize}
+
+Finally, incorporating regular expressions into the type system is not novel.
+The XDuce system \cite{pierce} typechecks XML schemas using regular expressions.
+We differ from this and related work in at least two ways.
+First, our system  is defined within an extensible type system;
+this first difference introduced an interesting class of design problems (see \S ???).
+TODO-nrf this is the second time this is used as a warrant. Factor out this argument
+and place it at the top???
+Second, our focus on security required novel design decisions; for instance, the
+filter elimination form is unique to Ace.
+
+In conclusion, our system is novel in at least two ways:
+\begin{itemize}
+\item The safety guarantees provided by libraries and frameworks in popular languages
+are not as (statically) justified as is often belived (or even claimed).
+\item Our extension is the first major demonstration of how an extensible type
+system may be used to provide light-weight, composable security analyses based upon
+idiomatic code.
+\end{itemize}
+
+\subsection{Outline}
+
+An outline of this paper follows:
+
+TODO-nrf real outline!
+\begin{itemize}
+  \item In \S 2, we define the type system's static and dynamic semantics.
+  \item Section 3 recalls some classical results about regular expressions and presents meta-theory for our system, including
+    the main soundness theorem for $\lcs$.
+  \item Finally, \S 4 discusses our implemention of $\lcs$ as a type system extension  within the Ace programming language.
+\end{itemize}
 
@@
 
-\section{Implementation Inside a Python}
+The $\lcs$ language is characterized by a type of strings indexed by regular
+expressions, together with operations on such strings which correspond to common
+input sanitation patterns.
+
+This section presents the grammar and semantics of $\lcs$.
+The semantics are defined in terms of an internal language with at least strings and a regex filter function.
+These constraints are captured by the internal term valuations ($\ival$).
+The internal language does not necessarily need a regex filter function because
+any dynamic conversion is easily definable using a combination of filters and safe
+casting.
+
+%
+% Begin Grammar 
+%
+\newcommand{\ereplace}[3]{{\tt replace}[#1](#2,#3)}
+\newcommand{\echeck}[2]{{\tt check}[#1](#2)}
+\newcommand{\estr}[2]{{\tt str[#1, #2]}}
+\newcommand{\estrt}[1]{{\tt strt[#1]}}
+\newcommand{\econcat}{+}
+
+\newcommand{\ireplace}[3]{{\tt ireplace}(#1,#2,#3)}
+\newcommand{\icheck}[2]{{\tt icheck}(#1, #2)}
+\newcommand{\istr}[1]{{{\tt istr}[#1]}}
+\newcommand{\istrt}{{\tt istrt}}
+\newcommand{\iconcat}{\hat{+}}
+
+% Special stuff for the semantics.
+\newcommand{\replacefn}[3]{\tt replacement(#1,#2,#3)}
+\newcommand{\crl}[3]{{\tt compute\_replacement\_lanugage}(#1,#2,#3)}
+
+\begin{figure}
+\begin{grammar}
+<r> ::= $\epsilon$ | $.$ | $a$ | $r \cdot r$ | $r + r$ | $r*$ \hfill Regex ($a \in \Sigma$).
+
+<$\tau$> ::=				\hfill	External types					\\
+$\tau \rightarrow \tau$ \alt
+$\estrt{r}$				\hfill Regular expression types
+
+<e> ::= \hfill External expressions \\
+      $x$ \alt
+      $\lambda x:\tau.e$ \alt
+      $e e$ \alt
+      $\estr{r}{s}$ \alt
+      $e \econcat e$ \hfill String concatenation \\
+      $\ereplace{r}{e}{e}$ \alt
+      $\echeck{r}{e}$
+
+<$\sigma$> ::= \hfill Internal types \\
+$\tau \rightarrow \tau$ \alt $\istrt$
+
+<i> ::= \hfill Internal expressions \\
+  $x$ \alt
+  $\lambda x:\sigma.i$ \alt
+  $i i$ \alt
+  $\istr{s}$ \alt
+  $i \econcat i$ \alt
+  $\ireplace{r}{i}{i}$ \alt
+  $\icheck{r}{i}$
+\end{grammar}
+\caption{Syntax of $\lcs$}
+\label{fig:lcsSyntax}
+\end{figure}
+
+%The $\lcs$ language gives static semantics for common regular expression library
+%functions. In this treatment, we include concatenation and filtering.
+%The ${\tt filter}$ function removes all instances of a regular expression in a string,
+%while concatenation ($+$) concatenates two strings.
+%
+%\subsection{Typing}
+%
+%The $\strin[r]$ type is parameterized by regular expressions; if $e:\strin[r]$,
+%then $e \in r$. Mapping from an arbitrary $\str$ to a $\strin[r]$ requires
+%defining an algorithm -- in terms of filter -- for converting a $\strin[.*]$
+%into a $\strin[r]$. The static semantics of the language defines the types of
+%operations on regular expressions in terms of well-understood properties about
+%regular lanugages; we recall these properties in section 3.
+%
+%\subsection{Dynamics}
+%
+%There are two evaluation judgements: $e:T \reduces e'$ and $e:T \ireduces i$.
+%The $\reduces$ relation is between $\lcs$ expressions, while the $\ireduces$
+%relation is a mapping from $\lcs$ expressions into internal language expressions
+%$i$ such that $i \ival$.
+%
+%Safety of the evaluation relation depends upon an injective mapping from $\lcs$ types info
+%internal language types. This relation, $h$, is defined below.
+%
+%\subsection{Type Safety}
+%
+%The type safety proof relies upon some assumptions about the type system and
+%dynamics of the internal language, as well as some properties of regular
+%languages.
+%
+%There must exist a translation from $\lcs$ types to the types of the internal
+%language. For the remainder of this paper, we call the type translation function
+%$h$.
+%
+%\begin{defn}[Type Translation Function $h$]
+%  The type translation function $h : Type \rightarrow IType$ is defined as follows:
+%  \begin{itemize}
+%    \item $\forall r. h(\strin[r]) = istr$
+%    \item $h(\str) = istr$
+%  \end{itemize}
+%\end{defn}
+%
+%Additionally, we assume that the internal language contains an implementation of strings,
+%together with operations for concatenation and filtering by regular expression.
+%
+%\begin{defn}[Types of internal values]
+%  Let $\istrf{s}$ range over string literals and $r$ over regular expressions.
+%  Internal values are typed as follows:
+%\begin{itemize}
+%  \item If $e=\istrf{s}$ then $e:istr$.
+%  \item If $e=\ifilter{r}{\istrf{s}}$ then $e:istr$.
+%  \item If $e=\istrf{s_1}+\istrf{s_2}$ then $e:istr$.
+%\end{itemize}
+%\end{defn}
+%
+%For simplicity, we assume a fixed translation from
+%$\lcs$ regular expressions to regular expressions recognizable by the internal
+%language's regex library (in practice, a fixed translation is acceptable.)
+%To summarize, we assume an internal language containing 
+%a string type together with operations for string concatentation and filtering. We
+%expect closure over strings for both operations.
+%Finally, recall that ${\tt ifilter}$ is only needed for dynamic casts, which may be removed without
+%descreasing the expressivity or even usability of the language.
+%Finally, the semantics of the filter function are defined in terms of ${\tt rl\_filter}$,
+%which is a static version of ${\tt ifilter}$.
+%
+\subsection{Properties of Regular Languages}
+
+The regular languages are the smallest set generated by regular expressions
+defined in Figure 1.
+
+\begin{thm}{Closure Properties.} \label{thm:closure}
+The regular expressions are closed under complements and concatenation.
+\end{thm}
+\begin{proof}
+See \cite{cinderella}.
+\end{proof}
+
+\begin{thm}{Coercion Theorem.} \label{thm:coerce}
+Suppose that $R$ and $L$ are regular expressions, and that $s \in R$ is a finite string.
+Let $s' := coerce(R,L,s)$ with all maximal substrings recognized by $L$ replaced with $\epsilon$.
+Then $s'$ is recognized by $(R \backslash L) + \emptyset$ and the construction of $R \backslash L$ is decidable.
+\end{thm}
+\begin{proof}
+Let $F,G$ be FAs corresponding to $R$ and $L$, and let $G'$ be $G$ with its final states inverted (so that $G'$ is the complement of $L$).
+Define an FA $H$ as a DFA corresponding to the NFA found by combining $F$ and $G'$ such that $H$ accepts only if $R$ and $L'$ accept or if $s$ is empty (this construction may result in an exponential blowup in state size.)  Clearly, $H$ corresponds to $R \backslash L + \emptyset$.
+Thus, the construction of $R \backslash L + \emptyset$ is decidable.
+
+If $R \subset L$, $s' = \emptyset$.
+If $L \subset R$, either $s' = \emptyset$, or $s' \in R$ and $s' \not \in L$.
+If $R$ and $L$ are not subsets of one another, then it may be the case that $L$ recognizes part of $R$.
+Consider $L$ as the union of two languages, one which is a subset of $R$ and one which is disjoint.
+The subset language is considered above and the disjoint language is inconsequential.
+\end{proof}
+
+%
+% Begin typing/translation relation
+%
+
+\onecolumn
+\begin{figure}
+\begin{center}
+\begin{tabular}{c r}
+
+\inferline{T-Var}
+{\gvd x : \tau}
+{x : \tau \in \Gamma}
+
+\inferline{T-Abs}
+{\gvd \lambda x : \tau .e : \tau \rightarrow \tau' \sreduces \lambda x : \sigma.i}
+{\Gamma, x : \tau \vdash e : \tau' \sreduces i \\ \sigma \equiv \trep(\tau)}
+
+\inferline{T-App}
+{\Gamma \vdash e_1 e_2 \sreduces i_1 i_2}
+{\Gamma \vdash e_1 : \tau \rightarrow \tau' \sreduces i_1 \\ \Gamma \vdash e_2 : \tau \sreduces i_2}
+
+\inferline{T-Replace}
+{\gvd \ereplace{r}{e_1}{e_2} : \estrt{r'} \sreduces \ireplace{r}{i_1}{i_2}}
+{\gvd e_1 : \estrt{r_1} \ireduces i_1 \\ \gvd e_2 : \estrt{r_2} \ireduces i_2 \\ r' \equiv \crl{r}{r_2}{r_1}}
+
+\inferline{T-Check-Ok}
+{\gvd \echeck{r'}{e} : \estrt{r'} \sreduces i}
+{\gvd e : \estrt{r} \sreduces{i} \\ \lang{r'} \subseteq \lang{r}}
+
+\inferline{T-Check-NotOk}
+{\gvd \echeck{r'}{e} : \estrt{r'} \sreduces \icheck{r}{i}}
+{\gvd e : \estrt{r} \sreduces i \\ \lang{r'} \not \subseteq \lang{r}}
+
+%\inferline{E-dconvert (optional) }
+%{\dconvert{\strin[r']}{e} : \strin[r'] \ireduces \ifilter{r'}{\istrf{e}}}
+%{e \val }
+
+\end{tabular}
+\caption{Type-directed translation to internal terms for $\lcs$}
+\label{fig:eeval}
+\end{center}
+\end{figure}
+
+%
+% Begin typing relation for internal terms.
+%
+\begin{figure}
+\begin{center}
+\begin{tabular}{c r}
+\inferline{Ti-Var}
+{\ovd x : \sigma}
+{x : \sigma \in \Omega}
+
+% Add in pedantic rules as well? If there's space.
+\inferline{Ti-Abs}
+{\ovd \lambda x : \sigma . i : \sigma \rightarrow \sigma'}
+{\Omega, x : \sigma \vdash i : \sigma'}
+
+\inferline{Ti-App}
+{i_1 i_2 : \sigma'}
+{\ovd i_1 : \sigma \rightarrow \sigma' \\ \ovd i_2 : \sigma}
+
+\inferline{Ti-Str}
+{\ovd \istr{s} : \istrt }
+{\cdot}
+
+\inferline{Ti-Replace}
+{\ovd \ireplace{r}{i_1}{i_2} : \istrt}
+{\ovd i_1 : \istrt \\ \ovd i_2 : \istrt}
+
+\inferline{Ti-Check}
+{\ovd \icheck{r}{i_1}{i_2} : \istrt}
+{\ovd i_1 : \istrt \\ \ovd i_2 : \istrt}
+
+\inferline{Ti-Concat}
+{\ovd i_1 \iconcat i_2 : \istrt}
+{\ovd i_1 : \istrt \\ \ovd i_2 : \istrt}
+
+\end{tabular}
+\caption{Typing relation on internal terms.}\label{fig:itypes}
+\end{center}
+\end{figure}
+
+%
+% Begin big-step sematnics for internal terms.
+%
+\newcommand{\err}{{\tt err}}
+\begin{figure}
+\begin{center}
+\begin{tabular}{c r}
+
+  \inferline{Ei-strval}
+  {\istr{s} \ival}
+  { \cdot }
+
+  \inferline{Ei-Replace}
+  {\ireplace{r}{i_1}{i_2} \ireduces i'}
+  { i \ireduces \replacefn{r}{i_1}{i_2} }
+
+  \inferline{Ei-CheckOk}
+  {\icheck{r}{i} \ireduces i}
+  {i \in \lang{r}}
+
+  \inferline{Ei-Check-NotOk}
+  {\icheck{r}{i} \ireduces \err}
+  {i \not \in \lang{r}}
+
+  \inferline{Ei-Concat}
+  { i_1 \iconcat i_2 \ireduces \istr{i_1 \iconcat i_2} }
+  { \cdot }
+
+\end{tabular}
+\caption{Big-step semantics for internal language; standard rules for the STL are omitted for brevity. The meta-variable $s$ ranges over string literals.}\label{fig:ieval}
+\end{center}
+\end{figure}
+
+\twocolumn
+
+Note that we never insert an internal check where the type of the string implies
+that a check must succeed. Furthermore, expensive calculations (such as language
+inclusion or ${\tt computer\_replacement\_language}$, all occur at compile time.
+Since sanitation problems are generally expressible with smalle, simple expressions,
+we believe that the compile-time overhead is not significant enough to prohibit
+our target usecase. Informal experimentation with our implementation seems to
+support this assertion.
+
+Due to Ei-Replace, all of our meta-theory depends upon a correct implementation
+of both regular expression replacement and inclusion checking in the internal language.
+We believe this assumption is okay for two reasons. First, our system is still
+superior to the status quo, which relies upon the correctness of these libraries
+\emph{in addition to} correct application logic. Second, regular expression libraries are
+generally well-tested and there exist verified implementations.
+
+\subsection{Type Safety Proof}
+
+\subsection{New Theorems}
+
+\begin{thm}[Progress for Internal Language]
+\end{thm}
+
+\begin{thm}[Preservation for Internal Language]
+\end{thm}
+
+\begin{thm}[Representational Consistency]
+If $\cdot \vdash e : \tau \sreduces i$ then $\cdot \vdash i : \sigma$ and $\sigma \equiv \trep(\tau)$.
+\end{thm}
+
+\begin{thm}[Main Soundness Result]
+  If $\cdot \vdash e : \estrt{r} \sreduces i$ then either $i \ireduces \err$ or
+  if $i \sreduces v$ for $v \ival$ then $v \in r$.
+\end{thm}
+
+%\begin{thm}[Preservation]
+%  Let $T$ be a type in $\lcs$ and $h(T)=\sigma$ the corresponding type in the internal language.
+%  For all terms $e$:
+%  \begin{itemize}
+%    \item If $e:T$ and $e:T \ireduces i$ then $i : \sigma$ such that $h(T) = \sigma$.
+%    \item If $e:T$ and $e:T \reduces e'$ then $e':T$.
+%  \end{itemize}
+%\end{thm}
+%\begin{proof}
+%The proof is a straightforward induction on the derivation of the combined evaluation relation.
+%\begin{itemize}[label=$ $,itemsep=1ex]
+%  \item \textbf{E-Ival, E-Ifilterval}. Both cases hold since the terms at hand are not $\lcs$ terms.
+%  \item \textbf{E-Stringval, E-strval}. Both cases hold since no reduction is possible. 
+%  \item \textbf{E-String}. By the definition of typing for internal terms, $\istrf{e} : istr$. It suffices to show that $h(\str) = istr$, which follows from the definition of $h$.
+%  \item \textbf{E-String\_in}.
+By the definition of typing for internal terms, $\istrf{e} : istr$.
+It suffices to show that $h(\strin[r]) = istr$, which follows from the definition of $h$ for arbitrary $r$.
+%  \item \textbf{E-concatval}.
+By the definition of typing for internal terms, $\istrf{e_1} + \istrf{e_2} : istr$.
+So it suffices to show that $h(\strin[r_1+r_2]) = istr$, which follows from the definition of $h$ for arbitrary $r_1$, $r_2$.
+%  \item \textbf{E-concatR, E-concatL}.
+Consider E-concatR.
+By induction, $e_1' : \strin{l_1}$.
+By inversion of T-Concat at the premise, $e_2 : \strin{r_2}$.
+Therefore, $e_1 + e_2 : \strin[r_1 + r_2]$.
+The left rule is symmetric.
+%  \item \textbf{E-Filterval}. We have that $\filter{\strin[r']}{e}:\strin[r\backslash r' + \emptyset]$. By inversion of T-Filter, $e:\strin[r]$.
+%    By T-Equiv-string\_in (which is bidirectional), $e \in \lang{r}$.
+%    By the Coercion Theorem, ${\tt rl\_filter}(r,e) \in \lang{r\backslash r' + \emptyset}$.
+%    By T-Equiv-string\_in,  $e \in \lang{r}$ and ${\tt rl\_filter}(r,e) \in \lang{r\backslash r' + \emptyset}$ implies ${\tt rl\_filter}(r,e):\strin[r\backslash r' + \emptyset]$.
+%  \item \textbf{E-Filter}. By inversion, $e : \strin[r']$ $\reduces e'$ so by the induction $e':\strin[r']$.
+%    Therefore, \\${\tt filter}(\strin[r], e'):$ $\strin[r \backslash r' + \emptyset]$ by T-Filter.
+%  \item \textbf{E-Convertval}. It suffices to show that $h(\strin[r]) = istr$, which is true by definition.
+%  \item \textbf{E-Convert}. By inversion and induction, $e' : \strin[r]$. We know that $[\strin[r']](e) : \strin[r']$, so by inversion of T-Convert $\lang{r'} \subseteq \lang{r}$.
+%    It follows that $[\strin[r']](e') : \strin[r']$.
+%  \item \textbf{E-DConvert}. By inversion, $e : \strin[r'] \reduces e'$. By the induction hypothesis, $e' : \strin[r']$; therefore, by T-Dconvert, $\dconvert[\strin[r]](e):\strin[r]$.
+%  \item \textbf{E-DConvertval}.
+By the definition of typing for internal terms, $\ifilter(r,\istrf{e}) : istr$.
+It suffices to show that $h(\strin[r_1+r_2]) = istr$, which follows from the definition of $h$.
+%\end{itemize}
+%\end{proof}
+%
+%\begin{thm}[Progress]
+%  If $e:T$ then $e:T \reduces^* e' \ireduces^* i$ where $i \ival$.
+%\end{thm}
+%\begin{proof}
+%By induction on the derivation of $e:T$.
+%For \textbf{T-Equiv-Include}, note that $e \in \lang{r}$, $e = \strf{s}$ for some $s$; therefore, $e \val$ by E-Strinval. 
+%We have that $e \val$ and $e : \strin[r]$, so $e \ireduces \istrf{e}$ by E-string\_in.
+%The remaining cases follow by induction on the hypotheses and application of corresponding evaluation rules.
+%\end{proof}
+%
+%\section{Implementation Inside a Python}
+%
+%TODO-nrf rewrite
+%The $\lcs$ language is implemented as a type system extension in Ace; this extension
+%is illustrated in the examples at the beginning of this paper.
+%
+%Computing a regular expression representating the language
+%$R \backslash S$ is necessary in order to type-check expressions in which the filter function
+%occurs. This language is computed by translating $R$ and $S$ into finite automata,
+%complementing the final states of $S$, then constructing the cross-product of $R$
+%and $S'$. Type checking terminates only because this construction is decidable.
+%
+%In addition to this construction, other more typical operations -- such
+%as equality checks for regular expressions and regular expression matching -- are
+%also necessary. For these reasons, the Ace extension implementing $\lcs$ includes a library
+%implementing each of these constructions.
+
 \subsection{Background: Ace}

### 2014-04-15 thesis-proposal/acc.tex

@@
 % !TEX root = omar-thesis-proposal.tex
+\lstset{tabsize=2, 
+basicstyle=\ttfamily\fontsize{8pt}{1em}\selectfont, 
+commentstyle=\itshape\ttfamily, 
+stringstyle=\ttfamily,
+numbers=left, numberstyle=\scriptsize\color{gray}\ttfamily, language=ML,moredelim=[il][\sffamily]{?},mathescape=true,showspaces=false,showstringspaces=false,xleftmargin=15pt, morekeywords=[1]{tyfam,opfam,let,fn,val,def,casetype,objtype,metadata,of,*,datatype,new,toast},deletekeywords={for,double,in},classoffset=0,belowskip=\smallskipamount,
+moredelim=**[is][\color{cyan}]{SSTR}{ESTR},
+moredelim=**[is][\color{OliveGreen}]{SHTML}{EHTML},
+moredelim=**[is][\color{purple}]{SCSS}{ECSS},
+moredelim=**[is][\color{brown}]{SSQL}{ESQL},
+moredelim=**[is][\color{orange}]{SCOLOR}{ECOLOR},
+moredelim=**[is][\color{magenta}]{SPCT}{EPCT}, 
+moredelim=**[is][\color{gray}]{SNAT}{ENDNAT}, 
+moredelim=**[is][\color{blue}]{SURL}{EURL},
+moredelim=**[is][\color{SeaGreen}]{SQT}{EQT},
+moredelim=**[is][\color{Periwinkle}]{SGRM}{EGRM},
+moredelim=**[is][\color{YellowGreen}]{SID}{EID},
+moredelim=**[is][\color{Sepia}]{SUS}{EUS}
+}
 \section{Active Code Completion}\label{acc}
-A language's syntax and semantics influences the behavior of a variety of tools beyond the compiler.
+An abstraction might benefit from support from a variety of tools beyond the compiler.
 For example, software developers today make heavy use of the code completion support found in modern source code editors  \cite{murphy_how_2006}.
-Editors for object-oriented languages provide code completion in the form of a floating menu containing  contextually-relevant variables, fields, methods, types and other code snippets.
+Editors for object-oriented languages provide code completion in the form of a floating menu containing  contextually-relevant variables, assignables, fields, methods, types and other code snippets.
 By navigating and selecting from this menu, developers are able to avoid many common spelling and logic errors, eliminate unnecessary keystrokes and explore unfamiliar APIs without incurring the mental overhead associated with switching to an external documentation tool.
@@
 
-\begin{lstlisting}
+\begin{lstlisting}[language=Java]
   public Color getDefaultColor() {
@@
 \includegraphics[width=40pc]{color_palette.png}\end{center}
-\caption{(a) An example code completion palette associated with the \texttt{Color} class. (b) The source code generated by this palette.}
+\caption{(a) An example code completion palette associated with the \texttt{Color} class. (b) The elaboration generated by this palette.}
 \label{colorpalette}
@@
 
-If the developer invokes the code completion command at the indicated cursor position (\li{_}), the editor  looks  for a {\it palette definition} associated with the {\it type} of the expression being entered, which in this case is  \verb|Color|.
+If the developer invokes the code completion command at the indicated cursor position (\li{_}), the editor  looks  for a {\it palette definition} associated with the {\it type} that the expression being entered is being analyzed against, which in this case is  \verb|Color| (due to the return type annotation on the method).
 If an associated palette is found, a menu item briefly describing this palette is added to the standard code completion menu.
@@
 The developer can interact with such palettes to provide parameters and other information related to their intent, and receive immediate feedback about the effect these choices will have on the behavior of the object being constructed.
-When this interaction is complete, the palette generates appropriate source code for insertion at the cursor.
-Figure \ref{colorpalette}(b) shows the inserted code after the user presses \textsc{enter}.
+When this interaction is complete, the palette produces an elaboration for insertion at the cursor, of the type (here, class) that the palette was associated with (Figure \ref{colorpalette}(b)).
 
+This is, in essence, an enriched edit-time variant of the mechanism used in Wyvern.
+As discussed in Sec.
+\ref{aparsing}, we associate a palette by providing metadata in the form of a class annotation.
+For example, we might write:
+\begin{lstlisting}[language=Java]
+@GraphitePalette(url="http://cs.cmu.edu/~comar/color-palette.html")
+class Color { ... }
+\end{lstlisting}
+
+Were a type-aware editor for Wyvern written, it might instead use metadata:
+\begin{lstlisting}
+objtype Color
+  ...
+  metadata : HasPalette = new
+    val palette_url = <SURLhttp://cs.cmu.edu/~comar/color-palette.htmlEURL>
+\end{lstlisting}
+
+\subsection{Design Process}
 We sought to address the following questions before designing and implementing our active code completion system:
@@
 \subsection{Remaining Tasks and Timeline}
-This work has been published at ICSE 2012 \cite{omar-graphite12}, and we do not plan on extending it further in this thesis.
-The main tasks have to do with relating it to the mechanisms in the previous sections, so that it generalizes beyond Java.
-We plan to do this when writing the thesis.
+This work has been published at ICSE 2012 \cite{Omar:2012:ACC:2337223.2337324}.
+The main remaining tasks have to do with specifying it in terms of the bidirectional typing mechanisms in Sec.
+3 more directly, rather than informally as in our paper.
+We plan to do this when writing the dissertation.
 There are also some pieces of related work that have been published since this paper was accepted that we need to review.

### 2014-04-15 thesis-proposal/background.tex

@@
 More powerful and direct compile-time \emph{term rewriting mechanisms} available in some languages can also be used for optimization, as well as for introducing specialized error checking logic and implementing new abstractions.
-For example, typed macros, such as those in MetaML \cite{Sheard:1999:UMS}, Template Haskell \cite{SheardPeytonJones:Haskell-02} and Scala \cite{scala-macros}, take full control over all of the code that they enclose.
+For example, typed macros, such as those in MetaML \cite{Sheard:1999:UMS}, Template Haskell \cite{SheardPeytonJones:Haskell-02} and Scala \cite{ScalaMacros2013}, take full control over all of the code that they enclose.
 This can be problematic, however, as outer macros can interfere with inner macros.

### 2014-04-15 thesis-proposal/parsing.tex

@@
 By defining abstractions in terms of existing general purpose constructs, we immediately know how to reason about them (here, by structural induction) and examine them (by pattern matching).
-They are already well optimized by compilers and benefit from editor support as well.
+They are already well optimized by compilers and benefit from general-purpose editor support as well.
 %The programmer only needs to provide an encoding of the structure of lists; the semantics and implementation are filled in by the general-purpose abstraction mechanism.
-While this is all quite useful, the associated general-purpose syntax is often not. For example, few would claim that writing a list of numbers as a sequence of \li{Cons} cells is convenient:
+While this is all quite useful, the associated general-purpose syntax is often less flexible.
+For example, few would claim that writing a list of numbers as a sequence of \li{Cons} cells is convenient:
 \begin{lstlisting}[numbers=none]
@@
 
-Because lists are a common data structure, many languages include \emph{literal notations} for introducing them, e.g.
+Because lists are a common data structure, many languages include \emph{literal notation} for introducing them, e.g.
 \li{[1, 2, 3, 4]}.
@@
 Using terminology from Green's cognitive dimensions of notations \cite{green1996usability}, it is more \emph{terse}, \emph{visible} and \emph{maps more closely} to the intuitive notion of a list.
-Stoy, in discussing the value of good notation, write \cite{stoy1977denotational}:
+Stoy, in discussing the value of good notation, writes \cite{stoy1977denotational}:
 \begin{quote}A good notation thus conceals much of the inner workings behind suitable abbreviations, while allowing us to consider it in more detail if we require: matrix and tensor notations provide further good examples of this.
-It may be summed up in the saying: 'A notation is important for what it leaves out.'\end{quote}
+It may be summed up in the saying: ``A notation is important for what it leaves out.''\end{quote}
 
 Although list, number and string literals are nearly ubiquitous features of modern languages, some languages also  provide specialized notation for other common data structures (like maps, sets, vectors and matrices), data formats (like XML and JSON), query languages (like regular expressions and SQL), markup languages (like HTML and Markdown) and many other types of data.
-For example, a language with built-in notation for HTML and SQL, supporting type-safe \emph{interpolation}\todo{splicing?} via curly braces, might define:
+For example, a language with built-in notation for HTML and SQL, supporting type-safe \emph{splicing} via curly braces, might define:
 \begin{lstlisting}
@@
 let webpage : HTML = HTMLElement(Dict.empty(), [BodyElement(Dict.empty(),
-  [H1Element(Dict.empty(), [TextNode(concat($\texttt{"}$SSTRResults for $\texttt{"}$ESTR, keyword))]), 
+  [H1Element(Dict.empty(), [TextNode($\texttt{"}$SSTRResults for $\texttt{"}$ESTR + keyword)]), 
   ULElement((Dict.add Dict.empty() ($\texttt{"}$SSTRid$\texttt{"}$ESTR, $\texttt{"}$SSTRresults$\texttt{"}$ESTR)), to_list_items(query(db, 
@@
 %%\end{lstlisting}
-This is problematic, for all of the reasons described above: it requires explicit conversions between representations, interference issues with string syntax, correctness problems, performance overhead and security issues.
+This is problematic, for all of the reasons described above: excessive conversions between representations, interference issues with string syntax, correctness problems, performance overhead and security issues.
 
@@
 These kinds of conflicts occur in a variety of circumstances: HTML and XML, different variants of SQL, JSON literals and dictionaries, or simply different implementations (``desugarings'') of the same specialized syntax (e.g.
-two regular expression engines) can all cause problems that may be difficult to anticipate.
-Techniques that limit the kinds of syntax extensions that can be expressed to guarantee ambiguities cannot occur simply cannot express these kinds of examples as-is (e.g.
+two regular expression engines).
+Techniques that limit the kinds of syntax extensions that can be expressed, to guarantee that ambiguities cannot occur, simply cannot express these kinds of examples as-is (e.g.
 \cite{conf/pldi/SchwerdfegerW09}).
@@
 This strategy avoids the problem of conflicting syntax, because neither the base language nor TSLs are ever extended directly.
-It also permits semantic flexibility -- the meaning of a form like \verb|{ }| can differ depending  on its type, so it is safe to use it for empty sets, maps and other data structures, like JSON literals,  freeing these common notations from being tied to the variant of a  data structure built into a language's standard library, which sometimes does not provide the exact semantics that a programmer needs (for example, Python dictionaries do not preserve key insertion order).
+It also permits semantic flexibility -- the meaning of a form like \verb|{ }| can differ depending  on its type, so it is safe to use it for empty sets, maps and other data structures, like JSON literals.
+This frees these common notations from being tied to the variant of a  data structure built into a language's standard library, which sometimes does not provide the exact semantics that a programmer needs (for example, Python dictionaries do not preserve key insertion order).
 \begin{figure}[t]
@@
 new : SearchServer
-  def resultsFor(searchQuery, page) =
+  def resultsFor(searchQuery, page)
     serve(~) (* serve : HTML -> Unit *)
@@
 \label{f-example}
-\vspace{-10px}
+%\vspace{-10px}
 \end{figure}
@@
 To allow us to focus on the essence of our proposal, the variant of Wyvern we will describe in this thesis is simpler than the variant previously described: it is purely functional (there are no effects other than non-termination) and it does not enforce a uniform access principle for objects (fields can be accessed directly).
-Objects can thus be thought of as recursive labeled products with support for with simple methods (functions that are automatically given a self-reference) for convenience.
+Objects can thus be thought of as recursive labeled products with support for simple methods (functions that are automatically given a self-reference) for convenience.
 We also add recursive labeled sum types, which we call \emph{case types}, that are quite similar to datatypes in ML.
@@
 \caption{Inline Generic Literal Forms}
-\vspace{-10px}
+%\vspace{-10px}
 \label{f-delims}
@@
 
-In addition to supporting conventional notation for URLs, this TSL supports \emph{interpolating}
-another Wyvern expression of type \li{URL} to form a larger URL. The interpolated term is delimited by percent signs,
-as seen on line 2 of Fig. \ref{f-example}. The TSL parses code between percent signs  as a Wyvern expression, using its AST to construct an AST for the expression as a whole. String concatenation  
-never occurs. Note that the delimiters used to go from Wyvern to a TSL are controlled by Wyvern  while the TSL controls how to return to Wyvern. 
+In addition to supporting conventional notation for URLs, this TSL supports \emph{splicing}
+another Wyvern expression of type \li{URL} to form a larger URL. The spliced term is delimited by percent signs,
+as seen on line 2 of Fig.
+\ref{f-example}.
+The TSL parses code between percent signs  as a Wyvern expression, using its abstract syntax tree (AST) to construct an AST for the expression as a whole.
+A string-based representation of the URL is never used at run-time.
+Note that the delimiters used to go from Wyvern to a TSL are controlled by Wyvern  while the TSL controls how to return to Wyvern.
 %\subsection{General-Purpose Notation for Object and Case Types}
@@
 The forward reference will be replaced by the general-purpose term, of type \li{HTML}, generated by the TSL during typechecking.
-The use of layout imposes no syntactic constraints on the body, while using an inline form does (Fig.
+Because layout was used as a delimiter, there are no syntactic constraints on the body, unlike with inline forms (Fig.
 \ref{f-delims}).
-For HTML, this is quite useful, as all of the inline forms impose constrains that would cause conflict with some valid piece of HTML.
+For HTML, this is quite useful, as all of the inline forms impose constrains that would cause conflict with some valid HTML.
 \subsection{Implementing a TSL}
@@
 casetype HTML 
-    Empty of Unit
-  | Text of String
-  | Seq of HTML * HTML 
-  | BodyElement of Attributes * HTML
-  | StyleElement  of Attributes * CSS
-  | ...
-  metadata = new : HasTSL
+  Empty
+  Seq of HTML * HTML 
+  Text of String
+  BodyElement of Attributes * HTML
+  StyleElement of Attributes * CSS
+  (* ... *)
+  metadata : HasTSL = new
     val parser = ~
-SGRM      start <- '>body'= start>
-        EGRMfn child => `SQTHTML.BodyElement(([], %EQTchildSQT%))EQT`
-SGRM      start <- '>style'= EXP>
-        EGRMfn e => `SQTHTML.StyleElement(([], %EQTeSQT%))EQT`
+SGRM      start <- '>body'= attributes start>
+        EGRMfn attrs, child => `SQTHTML.BodyElement((%EQTattrsSQT%, %EQTchildSQT%))EQT`
+SGRM      start <- '>style'= attributes EXP>
+        EGRMfn attrs, e => `SQTHTML.StyleElement((%EQTattrsSQT%, %EQTeSQT%))EQT`
 SGRM      start <- '<'= EXP>EGRM 
@@
 objtype Parser                          
-  def parse(ts : TokenStream) : (Exp * 
-    TokenStream)
-  metadata = new : HasTSL
-    val parser : Parser = (* parser generator *)      
-casetype Type
-    Var of ID
-  | Arrow of Type * Type
-  | Prod of Type * Type
-  metadata = new : HasTSL
-    val parser : Parser = (* type quasiquotes *)
+  def parse(ps : ParseStream) : ParseResult
+  metadata : HasTSL = new
+    val parser = (* parser generator *)
+casetype ParseResult
+  OK of Exp * ParseStream
+  Error of String * Location  
 \end{lstlisting}
@@
 \begin{subfigure}[t]{.42\textwidth}
-\begin{lstlisting}[linewidth=.42\textwidth,firstnumber=14]
+\begin{lstlisting}[linewidth=.42\textwidth,firstnumber=10]
 casetype Exp 
   Var of ID
-  | Lam of ID * Type * Exp
-  | Ap of Exp * Exp
-  | ProdIntro of Exp * Exp
-  | CaseIntro of Type * ID * Exp
+  Lam of ID * Type * Exp
+  Ap of Exp * Exp
+  Tuple of Exp * Exp
   ... 
-  | FromTS of Tokenstream
-  | Error of ErrorMessage
-  metadata = new : HasTSL
-    val parser : Parser = (* exp
-      quasiquotes *)
-...
+  Spliced of ParseStream
+  metadata : HasTSL = new
+    val parser = (* quasiquotes *)
 \end{lstlisting}
@@
 \caption{Some of the types included in the Wyvern prelude. They are mutually defined.}
-\vspace{-10px}
+%\vspace{-10px}
 \label{f-builtins}
@@
 For the purposes of this work, metadata values will always be of type \li{HasTSL}, an object type that declares a single field, \li{parser}, of type \li{Parser}.
-The \li{Parser} type is an object type declaring a single method, \li{parse}, that transforms a \li{TokenStream} extracted from a literal body to a Wyvern AST.
+The \li{Parser} type is an object type declaring a single method, \li{parse}, that transforms a \li{ParseStream} extracted from a literal body to a Wyvern AST.
 An AST is a value of type \li{Exp}, a case type that encodes the abstract syntax of Wyvern expressions.
@@
 \li{start}, or one of the built-in non-terminals \li{ID}, \li{EXP} or \li{TYPE}, representing Wyvern identifiers, expressions and types, respectively).
-Unique to Adams' formalism is that each terminal and non-terminal in a production can also have an optional \emph{layout constraint} associated with it.
+Unique to Adams grammars is that each terminal and non-terminal in a production can also have an optional \emph{layout constraint} associated with it.
 The layout constraints available are \li{=} (meaning that the leftmost column of the annotated term must be aligned with that of the parent term), \li{>} (the leftmost column must be indented further) and \li{>=} (the leftmost column \emph{may} be indented further).
@@
 \ref{f-builtins}).
-They support the full Wyvern concrete syntax as well as an additional delimited form, written with \li{%}s, that supports ``unquoting'': interpolating another AST into the one being generated.
-Again, interpolation is safe and structural, rather than based on string interpolation.
+They support the full Wyvern concrete syntax as well as an additional delimited form, written with \li{%}s, that supports ``unquoting'': splicing another AST into the one being generated.
+Again, splicing is safe and structural, rather than based on string interpolation.
 
-We have now seen several examples of TSLs that support interpolation.
-The question then arises: what type should the interpolated Wyvern expression be expected to have?
-This is determined by placing the interpolated value in a place in the generated AST where its type is known -- on line 11 of Fig.
+We have now seen several examples of TSLs that support splicing.
+The question then arises: what type should the spliced Wyvern expression be expected to have?
+This is determined by placing the spliced value in a place in the generated AST where its type is known -- on line 11 of Fig.
 \ref{f-htmltype} it is known to be \li{HTML} and on line 13 it is known to be \li{CSS} by the declaration of \li{HTML}, and on line 15, it is known to be \li{HTML} by the use of an explicit ascription.
@@
 \item We provide a complete layout-sensitive concrete syntax.
-We show how it can be written without the need for a context-sensitive lexer or parser using Adams' formalism and provide a full specification for the layout-delimited literal form introduced by a forward reference, \li{~}, as well as other forms of forward-referenced blocks (for the forms \li{new} and \li{case(e)}, in particular).
+We show how it can be written without the need for a context-sensitive lexer or parser using an Adams grammar and provide a full specification for the layout-delimited literal form as well as other forms of forward-referenced blocks (for the forms \li{new} and \li{case(e)}).
 %\item We detail the general mechanism for associating metadata with a type.
 A TSL is then implemented by associating a parser (of type \li{Parser}) with a type.
-The parser is responsible for rewriting tokenstreams (of type \li{Tokenstream}) into Wyvern ASTs (of type \li{Exp}).
+The parser is responsible for rewriting ParseStreams (of type \li{ParseStream}) into Wyvern ASTs (of type \li{Exp}).
 These types are defined in the standard library.
@@
 By distinguishing locations where an expression synthesizes a type from locations where an expression is being analyzed against a known type, we can precisely state where generic literals can and cannot appear and how parsing is delegated to a TSL.
-\item A na\"ive rewriting strategy would be \emph{unhygienic} -- it could allow for the inadvertent capture of local variables.
-We show a novel mechanism that ensures hygiene by requiring that the generated AST is closed except for subtrees derived from portions of the user's tokenstream that are interpreted as nested Wyvern expressions.
+The key judgements are of the form:
+\[\Gamma'; \Gamma \vdash_\Sigma e \leadsto \hat{e} \Leftarrow \tau~~~~~~~~\text{and}~~~~~~~~\Gamma'; \Gamma \vdash_\Sigma e  \leadsto \hat{e} \Rightarrow \tau\]
+
+Expressions, $e$, elaborate to \emph{literal-free expressions}, $\hat{e}$.
+This occurs simultaneously with typechecking; the first judgement captures situations where type analysis is occurring, the second type synthesis.
+The typing context, $\Gamma$, is standard, and the named type context, $\Sigma$, associates named types (object types and case types, but not arrow or product types) with their declarations and metadata.
+
+\item A na\"ive rewriting strategy would be \emph{unhygienic} -- it could allow for the inadvertent capture and shadowing of variables around a TSL literal.
+We show a novel mechanism that ensures hygiene by requiring that the generated AST is closed except for subtrees derived from portions of the user's parse stream that are interpreted as spliced Wyvern expressions or identifiers.
+We formalize this mechanism by splitting the context in our static semantics (in the judgement above, $\Gamma'$ contains variables only allowed in spliced Wyvern expressions).
 We also show how to explicitly refer to local values available in the parser definition (e.g.
 helper functions) in a safe way.
-We formalize this hygiene mechanism by splitting the context in our static semantics.
-\item To examine how broadly applicable the technique is, we conduct a simple corpus analysis, finding that string languages are used ubiquitously in existing Java code  (collaborative work with Darya Kurilova).
 \end{enumerate}
@@
 This work has been accepted to ECOOP 2014, having received quite positive reviews.
-Final revisions on the paper text are due on May 12th and we will defer these tasks to after the OOPSLA deadline in late March.% If not accepted, we will aim to resubmit to OOPSLA.
+Final revisions on the paper text are due on May 12th.% and we will defer these tasks to after the OOPSLA deadline in late March.% If not accepted, we will aim to resubmit to OOPSLA.
 \begin{enumerate}
@@
 We will add these simple rules to the grammar.
-\item We must write down the full formal semantics (including rules that we omitted for concision in the paper draft) in a technical report, as well as provide more complete proofs of the metatheory.
+\item There are a few presentation issues with our current formalism as written that need to be fixed.
+We must also write down the rules that we omitted for concision in the paper draft in a technical report, as well as provide more complete proofs of the metatheory.
 This might require slightly restricting the recursion in the rule for literals, so that induction can be shown to be well-founded.
@@
 \item We have not yet formalized our mechanism for preventing unintentional variable \emph{shadowing}, only unintentional variable \emph{capture}.
-It should be possible to prevent shadowing by preventing variables from leaking into interpolated Wyvern expressions (so that function application is the only way to pass data from a TSL to Wyvern code, as in the \li{Parser} TSL above).
+It should be possible to prevent shadowing by preventing variables from leaking into spliced Wyvern expressions (so that function application is the only way to pass data from a TSL to Wyvern code, as in the \li{Parser} TSL above).
 \item In macro systems like Scheme's, free variables in generated ASTs refer to their bindings within the macro definition, rather than where they are inserted.
 To support this, we have introduced the \li{toast} operator.
-Using it is currently explicit.
-We believe we can insert \li{toast} automatically using single variable (rather than whole expression) interpolation.
+Using it is currently explicit, so free variables that were not inside \li{toast} result in errors.
+We believe we can insert \li{toast} automatically using single variable (rather than whole expression) splicing so that free variables are implicitly \li{toast}ed, from within a library.
+We will show this.
+%\item To examine how broadly applicable the technique is, we conduct a simple corpus analysis, finding that string languages are used ubiquitously in existing Java code.
+We need to finish this (collaborative work with Darya Kurilova).
 \end{enumerate}
@@
 %- Figure out separate compilation (e.g. of CSS)
+\newpage

### 2014-04-15 sanitation-plas14/sanitation-plas14.tex
Beginnings of Sec 3 of PLAS paper
Removed <> from syntax
confusion

@@
-%TODO-nrf:
-% * Type system definition -- check big step semantics
-% PRoofs tonigt:
-% * Type safety proof for internal language
-% * Representational consistency
-
-%\documentclass[12pt]{article}
 \documentclass[10pt,preprint]{sigplanconf}
@@
 \newcommand{\ereplace}[3]{{\tt replace}[#1](#2,#3)}
-\newcommand{\echeck}[2]{{\tt check}[#1](#2)}
-\newcommand{\estr}[2]{{\tt str[#1, #2]}}
-\newcommand{\estrt}[1]{{\tt strt[#1]}}
-\newcommand{\econcat}{+}
+\newcommand{\ecoerce}[2]{{\tt coerce}[#1](#2)}
+\newcommand{\estr}[1]{{\tt str\_in[#1]}}
+\newcommand{\estrt}[1]{{\tt string\_in[#1]}}
+\newcommand{\econcat}[2]{{\tt concat}(#1,#2)}
 
@@
 \newcommand{\iconcat}{\hat{+}}
+\newcommand{\regext}{ {\tt regex} }
 
@@
 
+% Remove the < > from grammar productions.
+\renewcommand{\grammarlabel}[2]{#1\hfill#2}
+
 \begin{figure}
@@
 
-<$\tau$> ::=				\hfill	External types					\\
-$\tau \rightarrow \tau$ \alt
-$\estrt{r}$				\hfill Regular expression types
+<$\psi$> ::=				\hfill	External types					\\
+$\estrt{r}$				\hfill Regular expression types \alt
 
-<e> ::= \hfill External expressions \\
-      $x$ \alt
-      $\lambda x:\tau.e$ \alt
-      $e e$ \alt
-      $\estr{r}{s}$ \alt
-      $e \econcat e$ \hfill String concatenation \\
+<T> ::= \hfill External expressions \\
+      $\estr{s}$ \alt
+      $\econcat{e}{e}$ \hfill String concatenation \alt
       $\ereplace{r}{e}{e}$ \alt
-      $\echeck{r}{e}$
+      $\ecoerce{r}{e}$
 
-<$\sigma$> ::= \hfill Internal types \\
-$\tau \rightarrow \tau$ \alt $\istrt$
+<$\theta$> ::= \hfill Internal types \\
+$\istrt$ \alt $\regext$
 
-<i> ::= \hfill Internal expressions \\
-  $x$ \alt
-  $\lambda x:\sigma.i$ \alt
-  $i i$ \alt
+<S> ::= \hfill Internal expressions \\
   $\istr{s}$ \alt
-  $i \econcat i$ \alt
+  $\econcat{i}{i}$ \alt
   $\ireplace{r}{i}{i}$ \alt
@@
 \end{grammar}
-\caption{Syntax of $\lcs$}
+\caption{Syntax for the string sanitation fragment of $\lcs$}
 \label{fig:lcsSyntax}
@@
 \onecolumn
+\newcommand{\ectx}{?} % Context for external typing
+\newcommand{\ictx}{?} % Context for internal typing
+
+\newcommand{\ereduces}{\Downarrow}
+
 \begin{figure}
@@
 
-\inferline{T-Var}
-{\gvd x : \tau}
-{x : \tau \in \Gamma}
+\inferline{Te-include}
+{\ectx \vdash \estr{s} : \estrt{r}}
+{s \in \lang{r}???}
 
-\inferline{T-Abs}
-{\gvd \lambda x : \tau .e : \tau \rightarrow \tau' \sreduces \lambda x : \sigma.i}
-{\Gamma, x : \tau \vdash e : \tau' \sreduces i \\ \sigma \equiv \trep(\tau)}
+\inferline{Te-Concat}
+{\ectx \vdash \econcat{S_1}{S_2} : \estrt{r_1 \cdot r_2}}
+{\ectx \vdash S_1 : \estrt{r_1} \\ \ectx \vdash S_2 : \estrt{r_2}}
 
-\inferline{T-App}
-{\Gamma \vdash e_1 e_2 \sreduces i_1 i_2}
-{\Gamma \vdash e_1 : \tau \rightarrow \tau' \sreduces i_1 \\ \Gamma \vdash e_2 : \tau \sreduces i_2}
+\inferline{Te-subst}
+{\ectx \vdash \ereplace{r}{S_1}{S_2} : r}
+{\ectx \vdash S_1 : \estrt{r_1} \\ \ectx \vdash S_2 : \estrt{r_2} \\ r = {\tt lsubst}(r,r_1,r_2)}
 
-\inferline{T-Replace}
-{\gvd \ereplace{r}{e_1}{e_2} : \estrt{r'} \sreduces \ireplace{r}{i_1}{i_2}}
-{\gvd e_1 : \estrt{r_1} \ireduces i_1 \\ \gvd e_2 : \estrt{r_2} \ireduces i_2 \\ r' \equiv \crl{r}{r_2}{r_1}}
-
-\inferline{T-Check-Ok}
-{\gvd \echeck{r'}{e} : \estrt{r'} \sreduces i}
-{\gvd e : \estrt{r} \sreduces{i} \\ \lang{r'} \subseteq \lang{r}}
-
-\inferline{T-Check-NotOk}
-{\gvd \echeck{r'}{e} : \estrt{r'} \sreduces \icheck{r}{i}}
-{\gvd e : \estrt{r} \sreduces i \\ \lang{r'} \not \subseteq \lang{r}}
-
-%\inferline{E-dconvert (optional) }
-%{\dconvert{\strin[r']}{e} : \strin[r'] \ireduces \ifilter{r'}{\istrf{e}}}
-%{e \val }
+\inferline{Te-coerce}
+{\ectx \vdash \ecoerce{r'}{s} : \estrt{r'}}
+{\ectx \vdash S : \estrt{r}}
 
 \end{tabular}
-\caption{Type-directed translation to internal terms for $\lcs$}
+\caption{Typing rules for $\lcs$. Err cases?}
+\label{fig:etyping}
+\end{center}
+\end{figure}
+
+\begin{figure}
+\begin{center}
+\begin{tabular}{c r}
+\inferline{Te-Include}
+{\estr{s} \val}
+{ \cdot }
+
+\inferline{Te-Concat}
+{\econcat{S_1}{S_2} \ereduces \estr{s_1 \cdot s_2}} % ???
+{S_1 \Downarrow \estr{s_1} \\ S_2 \Downarrow \estr{s_2}}
+
+\inferline{Te-Subst}
+{\ereplace{r}{S_1}{S_2} \ereduces s}
+{S_1 \Downarrow \estr{s_1} \\ S_2 \Downarrow \estr{s_2} \\ s = {\tt substitution}(r,s_1,s_2)}
+
+\inferline{Te-Coerce-OK}
+{\ecoerce{r}{s} \ereduces \estr{x}}
+{s \ereduces \estr{x} \\ x \in \lang{r}}
+
+\inferline{Te-Coerce-NotOk}
+{\ecoerce{r}{s} \ereduces \estr{x} \\ x \not \in \lang{r}}
+{s \ereduces \estr{x} \\ x \not \in \lang{r}}
+\end{tabular}
+\caption{Operational semantics for the external language $\lcs$.
+There are also unmentioned error propagation rules for for Te-Concat and Te-Subst.}
 \label{fig:eeval}
@@
 \end{figure}
+
+\begin{figure}
+\begin{center}
+\begin{tabular}{c r}
+\inferline{Ti-Include}
+{\istr{s} \val}
+{ \cdot }
+
+\inferline{Ti-Include}
+{{\tt regex}[r] \val}
+{ \cdot }
+
+\inferline{Ti-Concat}
+{\iconcat{S_1}{S_2} \ireduces \istr{s_1 \cdot s_2}} % ???
+{S_1 \ireduces \istr{s_1} \\ S_2 \ireduces \istr{s_2}}
+
+\inferline{Te-Subst}
+{\ireplace{r}{S_2}{S_3} \ireduces {\tt substitution(r,s_2,s_3)}}
+{T_1 \ireduces {\tt regex}[r] \\ T_2 \ireduces \istr{s_2} \\ T_3 \ireduces \istr{s_3} \\ }
+
+\inferline{Te-Ccheck-OK}
+{\icheck{r}{s} \ireduces \istr{x}}
+{s \ireduces \estr{x} \\ x \in \lang{r}}
+
+\inferline{Te-Check-NotOk}
+{\icheck{r}{s} \ireduces \istr{x} \\ x \not \in \lang{r}}
+{s \ireduces \istr{x} \\ x \not \in \lang{r}}
+\end{tabular}
+\caption{Operational semantics for the internal language of $\lcs$.
+There are also unmentioned error propagation rules for for Te-Concat and Te-Subst.}
+\label{fig:ieval}
+\end{center}
+\end{figure}
+
+%
+%\begin{figure}
+%\begin{center}
+%\begin{tabular}{c r}
+%
+%\inferline{T-Abs}
+%{\gvd \lambda x : \tau .e : \tau \rightarrow \tau' \sreduces \lambda x : \sigma.i}
+%{\Gamma, x : \tau \vdash e : \tau' \sreduces i \\ \sigma \equiv \trep(\tau)}
+%
+%\inferline{T-App}
+%{\Gamma \vdash e_1 e_2 \sreduces i_1 i_2}
+%{\Gamma \vdash e_1 : \tau \rightarrow \tau' \sreduces i_1 \\ \Gamma \vdash e_2 : \tau \sreduces i_2}
+%
+%\inferline{T-Replace}
+%{\gvd \ereplace{r}{e_1}{e_2} : \estrt{r'} \sreduces \ireplace{r}{i_1}{i_2}}
+%{\gvd e_1 : \estrt{r_1} \ireduces i_1 \\ \gvd e_2 : \estrt{r_2} \ireduces i_2 \\ r' \equiv \crl{r}{r_2}{r_1}}
+%
+%\inferline{T-Check-Ok}
+%{\gvd \echeck{r'}{e} : \estrt{r'} \sreduces i}
+%{\gvd e : \estrt{r} \sreduces{i} \\ \lang{r'} \subseteq \lang{r}}
+%
+%\inferline{T-Check-NotOk}
+%{\gvd \echeck{r'}{e} : \estrt{r'} \sreduces \icheck{r}{i}}
+%{\gvd e : \estrt{r} \sreduces i \\ \lang{r'} \not \subseteq \lang{r}}
+%  
+%%\inferline{E-dconvert (optional) }
+%%{\dconvert{\strin[r']}{e} : \strin[r'] \ireduces \ifilter{r'}{\istrf{e}}}
+%%{e \val }
+%
+%\end{tabular}
+%\caption{Type-directed translation to internal terms for $\lcs$}
+%\label{fig:eeval}
+%\end{center}
+%\end{figure}
 
@@
 %
-%\section{Implementation Inside a Python}
+\section{Implementation Inside a Python}
 %
@@
 %implementing each of these constructions.
+\newcommand{\F}[1]{\textsf{#1}~}
+\newcommand{\FF}[1]{\textsf{#1}}
+\newcommand{\Q}{\FF{Arg}}
+\newcommand{\xlA}[1]{\lfloor #1 \rfloor_{\lamAce}}
+\newcommand{\xA}[1]{$\lfloor #1 \rfloor_{\texttt{ace}}$}
+\begin{figure*}
+\begin{tabular}{ l r l l }
+$\tau$ & & $\xlA{\tau}$ & \xA{\tau}\\
+\hline
+$\estrt{r}$ & & $\fvar{stringin}[r]$ & \verb|string_in["|$r$\verb|"]|\\
+\\
+  $e$ & & $\xlA{e}$ & \xA{e}\\
+  \hline
+  ${\tt str}[r, s]$ & \text{synthetic position} & $\FF{intro}[\FF{str}[s]]()$ : \fvar{stringin}[\FF{rx}[$r$]] & \verb|"|$s$\verb|" (string_in["|$r$\verb|"])|\\
+  & \text{analytic position} & $\FF{intro}[\FF{str}[s]]()$ & \verb|"|$s$\verb|"|\\
+  ${\tt concat}(e_1; e_2)$ & & $\xlA{e_1}\cdot\FF{elim}[\tvar{concat}](\xlA{e_2})$ & \xA{e_1}\verb| + |\xA{e_2} \\
+  ${\tt replace}[r](e_1; e_2)$ & & $\xlA{e_1}\cdot\FF{elim}[\tvar{replace}~\FF{rx}[r]](\xlA{e_2})$ & \xA{e_1}\verb|.replace("|$r$\verb|", |\xA{e_2}\verb|)|\\
+  ${\tt coerce}[r](e)$ & & $\xlA{e}\cdot\FF{elim}[\tvar{coerce}~\FF{rx}[r]]()$ & \xA{e}\verb|.coerce("|$r$\verb|")|
+\end{tabular}
+\caption{Translation of $\lcs$ to $\lamAce$ and Ace.}
+\end{figure*}
+\begin{figure}
+\small\begin{flalign}
+& \F{tycon}\fvar{stringin}~\F{of}\FF{R}~\{\\
+& \quad \F{iana}\{\tlam{opidx}{\FF{String}}{
+	\tlam{tyidx}{\FF{R}}{
+	\tlam{a}{\klist{\Q}}{\\
+& \quad\quad \tvar{arity0}~\tvar{a}~(\tvar{check}~\tvar{opidx}~\tvar{tyidx}~\titerm{\iup{\tvar{opidx}}})
+	}}
+}\}\\
+& \quad \F{esyn}\{\tlam{opidx}{\kunit+(\FF{R}+\FF{R})}{
+	\tlam{a}{\klist{\Q}}{\\
+& \quad\quad \tsumcase{\tvar{opidx}}{\_}{\\
+& \quad\quad\quad \tvar{arity2}~\tvar{a}~\tlam{a1}{\Q}{\tlam{a2}{\Q}{\\
+& \quad\quad\quad\quad \tvar{rsyn}~\tvar{a1}~\tlam{r1}{\FF{R}}{\tlam{i1}{\kITerm}{~}}\\
+& \quad\quad\quad\quad \tvar{rsyn}~\tvar{a2}~\tlam{r2}{\FF{R}}{\tlam{i2}{\kITerm}{~}}\\
+& \quad\quad\quad\quad\quad (\ttype{stringin}{\FF{rseq}(\tvar{r1}; \tvar{r2})}, \\
+& \quad\quad\quad\quad\quad\titerm{{\tt iconcat}(\iup{\tvar{i1}}; \iup{\tvar{i2}})})
+}}\\
+& \quad\quad}{opidx'}{d}}
+}\}
+\}
+\end{flalign}
+\caption{Definition of ...}
+\end{figure}
+\begin{figure}
+\[
+\begin{array}{lcl}
+% & & Definition & Kind\\
+\tvar{concat} & := & \tinl{\kunit, R+R}{b}\\%& \kunit + (R + R)\\
+\tvar{replace} & := & \tlam{r}{R}{\tinr{\kunit, R+R}{\tinl{R,R}{\tvar{r}}}}\\% & \karrow{a}{b}\\
+\tvar{coerce} & := & \tlam{r}{R}{\tinr{\kunit, R+R}{\tinr{R,R}{\tvar{r}}}}\\
+\end{array}
+\]
+\caption{Definitions}
+\end{figure}
 

### 2014-04-16 sanitation-plas14/sanitation-plas14.tex
S. 2 figures converging on final version.

@@
     \makeatletter
-%
+
 % \btIfInRange{number}{range list}{TRUE}{FALSE}
@@
 \newcommand{\Lagr}{\mathcal{L}}
-\newcommand{\strin}{{\tt string\_in}}
 \newcommand{\lang}[1]{\Lagr\{#1\}}
 \newcommand{\stru}[2]{ {\tt string\_union}(#1,#2)}
-
-\newcommand{\str}{ {\tt string} }
 
@@
 \newcommand{\reduces}{ \Rightarrow }
-\newcommand{\ireduces}{ \Downarrow }
-\newcommand{\sreduces}{ \leadsto }
 \newcommand{\gvd}{\Gamma \vdash }
@@
 
-\newcommand{\istrf}[1]{`#1\textrm'} %??
+\newcommand{\tstrf}[1]{`#1\textrm'} %??
 \newcommand{\strf}[1]{``#1"}
 
-%
-% Constrain the size of full-page diagrams and rule lists
-%
-%\newcommand{\pagewidth}{5in}
-%\newcommand{\rulelistwidth}{3in}
+%%
+%% Source and Target language definitions.
+%%
+% Source language terms.
+\newcommand{\sisubst}[3]{{\tt subst}[#1](#2,#3)}
+\newcommand{\coerce}[2]{ {\tt coerce}[#1](#2)}
+\newcommand{\sistr}[1]{{\tt si\_str}[#1]}
+\newcommand{\strin}[1]{\sistr{#1}}
+\newcommand{\siconcat}[2]{{\tt siconcat}(#1,#2)}
 
-%
+% Source language types.
+\newcommand{\stringin}[1]{{\tt string\_in}[#1]}
+
+% target language terms.
+\newcommand{\tsubst}[3]{{\tt tsubst}(#1,#2,#3)}
+\newcommand{\tcheck}[2]{{\tt tcheck}(#1, #2)}
+\renewcommand{\tstr}[1]{{{\tt tstr}[#1]}}
+\newcommand{\tconcat}[2]{{\tt tconcat}(#1,#2)}
+
+% Target language types
+\newcommand{\str}{{\tt string}}
+\newcommand{\regext}[1]{ {\tt rx}[#1] }
+\newcommand{\rx}[1]{ {\tt rx}[#1] }
+
+% Meta-theoretic functions
+\newcommand{\dosubst}[3]{{\tt dosubst}(#1,#2,#3)}
+\newcommand{\lsubst}[3]{{\tt lsubst}(#1,#2,#3)}
+
+% Judgements
+\newcommand{\err}{{\tt err}}
+\newcommand{\trden}[1]{[[#1]]} % = Translation Denotation.
+
+% Relations
+\newcommand{\treduces}{ \Downarrow }
+\newcommand{\sreduces}{ \Downarrow }
+
+%%
+%% Constrain the size of full-page diagrams and rule lists
+%%
+%%\newcommand{\pagewidth}{5in}
+%%\newcommand{\rulelistwidth}{3in}
+
 % Names of type systems presented in paper
-%
 \newcommand{\lcs}{\lambda_{CS}}
@@
 \setlength{\grammarindent}{3em}
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 
@@
 
+% Whatever just dropping this in somewhere random.
+\begin{thm}
+If $S : \strin{r}$ then there exists a T such that $\trden{s} = T$ and either: 
+\begin{itemize}
+\item (a) $T \treduces \tstr{s}$ and $S \sreduces \strin{s}$, and $s \in lang{r}$.
+\item (b) $T \err$ and $S \err$.
+\end{itemize}
+\end{thm}
+\begin{proof}
+By induction on the derivation of $\trden{S} = T$. TODO-nrf add case magic.
+\begin{itemize}
+\item Tr-string. By assumption $\trden{S} = \tstr{s}$. The (a) property follows.
+By S-E-Str, $S \sreduces \str{s}$ and by T-E-Str, $\tstr{s} \treduces \tstr{s}$.
+By source language type safety, it follows that $s \in \lang{r}$.
+\item Tr-Concat.
+By inversion of the typing relaiton for the source lnaguage, $S_1 : \strin{r_1}$ and $S_2 : \stringin{S_2}$.
+By induction on the left premise, $T_1 \treduces \tstr{s_1}$, $S_1 \sreduces \str{s_1}$ and $s_1 \in \lang{r_1}$
+or else $T_1 err$. 
+Similarly for $T_2$ and $S_2$.
+In the error cases, the propagations rules corresponding to S-E-Concat and T-E-Concat imply that $\tconcat{T_1}{T_2} \err$ and $\siconcat{S_1}{S_2} \err$.
+In the non-error case, $\siconcat{S_1}{S_2} \sreduces \strin{s_1 \cdot s_2}$ by S-E-Concat and $\tconcat{T_1}{T_2} \treduces \tstr{s_1 \cdot s_2}$ by T-E-Concat.
+Finally, $s_1 \in \lang{r_1}$ and $s_2 \in \lang{r_2}$ implies that $s_1 \cdot s_2 \in \lang{r}$ by \ref{lemmaregex}.
+\item T-subst.
+Suppose $\sisubst{r}{S_1}{S_2} : \strin{s}$ for some $x$. 
+By inverison, $S_1 : \strin{r_1}$ and $S_2 : \strin{r_2}$.
+Note that $s : \subst{r}{S_1}{S_2} : \stringin{\tsubst{r}{r_1}{r_2}}$ by S-T-Subst.
+By induction, $\trden{S_1} = T_1$ and $\trden{S_2} = T_2$.
+Either (a) or (b) holds for each.
+
+If (a) holds for both, the $S_1 \sreduces \strin{S_1}$ and $S_2 \sreduces \strin{S_2}$ for $s_1 \in \lang{r_1}$ and $s_2 \in \lang{r_2}$.
+Therefore, $\sisubst{r}{S_1}{S_2} \sreduces \strin{\dosubst{r,s_1,s_2}}$.
+The same fact can be shown in an analogous manner for $T_1$ and $T_2$.
+Finally, $s_1 \in \lang{r_1}$ and $s_2 \in \lang{r_2}$, which implies that $\dosubst{r}{s_1}{s_2} \in \lang{\sisubst{r}{s_1}{s_2}}$ by \ref{substhm}.
+
+If (b) holds for either the left or right premise, then $\sisubst{r}{S_1}{S_2} \err$ by the propagation rules corresponding to S-T-subst, 
+and $\tsubst{S_1}{S_2} \err$ by the propagation rules corresponding to T-T-Subst.
+\end{itemize}
+\end{proof}
+
 %
@@
 %
-\newcommand{\ereplace}[3]{{\tt replace}[#1](#2,#3)}
-\newcommand{\ecoerce}[2]{{\tt coerce}[#1](#2)}
-\newcommand{\estr}[1]{{\tt str\_in[#1]}}
-\newcommand{\estrt}[1]{{\tt string\_in[#1]}}
-\newcommand{\econcat}[2]{{\tt concat}(#1,#2)}
-
-\newcommand{\ireplace}[3]{{\tt ireplace}(#1,#2,#3)}
-\newcommand{\icheck}[2]{{\tt icheck}(#1, #2)}
-\newcommand{\istr}[1]{{{\tt istr}[#1]}}
-\newcommand{\istrt}{{\tt istrt}}
-\newcommand{\iconcat}{\hat{+}}
-\newcommand{\regext}{ {\tt regex} }
-
-% Special stuff for the semantics.
-\newcommand{\replacefn}[3]{\tt replacement(#1,#2,#3)}
-\newcommand{\crl}[3]{{\tt compute\_replacement\_lanugage}(#1,#2,#3)}
 
@@
 
-<$\psi$> ::=				\hfill	External types					\\
-$\estrt{r}$				\hfill Regular expression types \alt
+<$\psi$> ::=				\hfill	Source language types					\\
+$\stringin{r}$				\hfill Regular expression types \alt
 
-<T> ::= \hfill External expressions \\
-      $\estr{s}$ \alt
-      $\econcat{e}{e}$ \hfill String concatenation \alt
-      $\ereplace{r}{e}{e}$ \alt
-      $\ecoerce{r}{e}$
+<S> ::= \hfill Source language expressions \\
+      $\strin{S}$ \alt
+      $\siconcat{S}{S}$ \hfill String concatenation \alt
+      $\sisubst{r}{S}{S}$ \alt
+      $\coerce{r}{S}$
 
-<$\theta$> ::= \hfill Internal types \\
-$\istrt$ \alt $\regext$
+<$\theta$> ::= \hfill Target language types \\
+$\str$ \alt $\regext$
 
-<S> ::= \hfill Internal expressions \\
-  $\istr{s}$ \alt
-  $\econcat{i}{i}$ \alt
-  $\ireplace{r}{i}{i}$ \alt
-  $\icheck{r}{i}$
+<T> ::= \hfill Internal expressions \\
+  $\tstr{T}$ \alt
+  $\siconcat{T}{T}$ \alt
+  $\tsubst{r}{T}{T}$ \alt
+  $\tcheck{r}{T}$
 \end{grammar}
@@
 %
-%There are two evaluation judgements: $e:T \reduces e'$ and $e:T \ireduces i$.
-%The $\reduces$ relation is between $\lcs$ expressions, while the $\ireduces$
+%There are two evaluation judgements: $e:T \reduces e'$ and $e:T \treduces i$.
+%The $\reduces$ relation is between $\lcs$ expressions, while the $\treduces$
 %relation is a mapping from $\lcs$ expressions into internal language expressions
@@
 %\begin{defn}[Types of internal values]
-%  Let $\istrf{s}$ range over string literals and $r$ over regular expressions.
+%  Let $\tstrf{s}$ range over string literals and $r$ over regular expressions.
 %  Internal values are typed as follows:
 %\begin{itemize}
-%  \item If $e=\istrf{s}$ then $e:istr$.
-%  \item If $e=\ifilter{r}{\istrf{s}}$ then $e:istr$.
-%  \item If $e=\istrf{s_1}+\istrf{s_2}$ then $e:istr$.
+%  \item If $e=\tstrf{s}$ then $e:istr$.
+%  \item If $e=\ifilter{r}{\tstrf{s}}$ then $e:istr$.
+%  \item If $e=\tstrf{s_1}+\tstrf{s_2}$ then $e:istr$.
 %\end{itemize}
@@
 \onecolumn
-\newcommand{\ectx}{?} % Context for external typing
-\newcommand{\ictx}{?} % Context for internal typing
+\newcommand{\sctx}{\Gamma} % Context for external typing
+\newcommand{\tctx}{\Omega} % Context for internal typing
 
@@
 
-\inferline{Te-include}
-{\ectx \vdash \estr{s} : \estrt{r}}
-{s \in \lang{r}???}
+\inferline{S-T-include}
+{\sctx \vdash \strin{s} : \stringin{r}}
+{s \in \lang{r}}
 
-\inferline{Te-Concat}
-{\ectx \vdash \econcat{S_1}{S_2} : \estrt{r_1 \cdot r_2}}
-{\ectx \vdash S_1 : \estrt{r_1} \\ \ectx \vdash S_2 : \estrt{r_2}}
+\inferline{S-T-Concat}
+{\sctx \vdash \siconcat{S_1}{S_2} : \strin{r_1 \cdot r_2}}
+{\sctx \vdash S_1 : \strin{r_1} \\ \sctx \vdash S_2 : \strin{r_2}}
 
-\inferline{Te-subst}
-{\ectx \vdash \ereplace{r}{S_1}{S_2} : r}
-{\ectx \vdash S_1 : \estrt{r_1} \\ \ectx \vdash S_2 : \estrt{r_2} \\ r = {\tt lsubst}(r,r_1,r_2)}
+\inferline{S-T-subst}
+{\sctx \vdash \sisubst{r}{S_1}{S_2} : r'}
+{\sctx \vdash S_1 : \strin{r_1} \\ \sctx \vdash S_2 : \strin{r_2} \\ {\tt lsubst}(r,r_1,r_2) = r'}
 
-\inferline{Te-coerce}
-{\ectx \vdash \ecoerce{r'}{s} : \estrt{r'}}
-{\ectx \vdash S : \estrt{r}}
+\inferline{S-T-coerce}
+{\sctx \vdash \coerce{r'}{s} : \stringin{r'}}
+{\sctx \vdash S : \stringin{r}}
 
 \end{tabular}
-\caption{Typing rules for $\lcs$. Err cases?}
+\caption{Typing rules for source langauge terms (S). The metavariable s ranges over
+string literals.}
 \label{fig:etyping}
@@
 \inferline{Te-Include}
-{\estr{s} \val}
+{\strin{s} \val}
 { \cdot }
 
-\inferline{Te-Concat}
-{\econcat{S_1}{S_2} \ereduces \estr{s_1 \cdot s_2}} % ???
-{S_1 \Downarrow \estr{s_1} \\ S_2 \Downarrow \estr{s_2}}
+\inferline{S-E-Concat}
+{\siconcat{S_1}{S_2} \ereduces \strin{s_1 \cdot s_2}} % ???
+{S_1 \Downarrow \strin{s_1} \\ S_2 \Downarrow \strin{s_2}}
 
-\inferline{Te-Subst}
-{\ereplace{r}{S_1}{S_2} \ereduces s}
-{S_1 \Downarrow \estr{s_1} \\ S_2 \Downarrow \estr{s_2} \\ s = {\tt substitution}(r,s_1,s_2)}
+\inferline{S-E-Subst}
+{\sisubst{r}{S_1}{S_2} \ereduces s}
+{S_1 \Downarrow \strin{s_1} \\ S_2 \Downarrow \strin{s_2} \\ \dosubst{r}{s_1}{s_2} = s}
 
-\inferline{Te-Coerce-OK}
-{\ecoerce{r}{s} \ereduces \estr{x}}
-{s \ereduces \estr{x} \\ x \in \lang{r}}
+\inferline{S-E-Coerce-OK}
+{\coerce{r}{S} \ereduces \strin{s}}
+{S \ereduces \strin{s} \\ x \in \lang{r}}
 
-\inferline{Te-Coerce-NotOk}
-{\ecoerce{r}{s} \ereduces \estr{x} \\ x \not \in \lang{r}}
-{s \ereduces \estr{x} \\ x \not \in \lang{r}}
+\inferline{S-E-Coerce-NotOk}
+{S \err}
+{\coerce{r}{S} \ereduces \strin{s} \\ s \not \in \lang{r}}
+
 \end{tabular}
-\caption{Operational semantics for the external language $\lcs$.
-There are also unmentioned error propagation rules for for Te-Concat and Te-Subst.}
+\caption{Big step semantics for the string fragment of the source language.
+There are also unmentioned error propagation rules for for S-E-Concat and S-E-Subst.}
 \label{fig:eeval}
@@
 \begin{tabular}{c r}
-\inferline{Ti-Include}
-{\istr{s} \val}
-{ \cdot }
 
-\inferline{Ti-Include}
-{{\tt regex}[r] \val}
-{ \cdot }
+\inferline{T-T-String}
+{\tctx \vdash \tstr{s} : \str}
+{\cdot}
 
-\inferline{Ti-Concat}
-{\iconcat{S_1}{S_2} \ireduces \istr{s_1 \cdot s_2}} % ???
-{S_1 \ireduces \istr{s_1} \\ S_2 \ireduces \istr{s_2}}
+\inferline{T-T-Concat}
+{\tctx \vdash \tconcat{T_1}{T_2} : \str}
+{\tctx \vdash T_1 : \str \\ \tctx \vdash T_2 : \str}
 
-\inferline{Te-Subst}
-{\ireplace{r}{S_2}{S_3} \ireduces {\tt substitution(r,s_2,s_3)}}
-{T_1 \ireduces {\tt regex}[r] \\ T_2 \ireduces \istr{s_2} \\ T_3 \ireduces \istr{s_3} \\ }
+\inferline{T-T-Subst}
+{\tctx \vdash \tsubst{\rx{r}}{T_1}{T_2} : \str}
+{\tctx \vdash T_1 : \str \\ \tctx \vdash T_2 : \str}
 
-\inferline{Te-Ccheck-OK}
-{\icheck{r}{s} \ireduces \istr{x}}
-{s \ireduces \estr{x} \\ x \in \lang{r}}
+\inferline{T-T-Check}
+{\tctx \vdash \tcheck{\rx{r}}{s} : \str}
+{\tctx \vdash T : \str}
 
-\inferline{Te-Check-NotOk}
-{\icheck{r}{s} \ireduces \istr{x} \\ x \not \in \lang{r}}
-{s \ireduces \istr{x} \\ x \not \in \lang{r}}
 \end{tabular}
-\caption{Operational semantics for the internal language of $\lcs$.
-There are also unmentioned error propagation rules for for Te-Concat and Te-Subst.}
-\label{fig:ieval}
+\caption{Typing rules for target langauge terms (T). The metavariable s ranges over
+string literals.}
+\label{fig:etyping}
 \end{center}
@@
 
-%
-%\begin{figure}
-%\begin{center}
-%\begin{tabular}{c r}
-%
-%\inferline{T-Abs}
-%{\gvd \lambda x : \tau .e : \tau \rightarrow \tau' \sreduces \lambda x : \sigma.i}
-%{\Gamma, x : \tau \vdash e : \tau' \sreduces i \\ \sigma \equiv \trep(\tau)}
-%
-%\inferline{T-App}
-%{\Gamma \vdash e_1 e_2 \sreduces i_1 i_2}
-%{\Gamma \vdash e_1 : \tau \rightarrow \tau' \sreduces i_1 \\ \Gamma \vdash e_2 : \tau \sreduces i_2}
-%
-%\inferline{T-Replace}
-%{\gvd \ereplace{r}{e_1}{e_2} : \estrt{r'} \sreduces \ireplace{r}{i_1}{i_2}}
-%{\gvd e_1 : \estrt{r_1} \ireduces i_1 \\ \gvd e_2 : \estrt{r_2} \ireduces i_2 \\ r' \equiv \crl{r}{r_2}{r_1}}
-%
-%\inferline{T-Check-Ok}
-%{\gvd \echeck{r'}{e} : \estrt{r'} \sreduces i}
-%{\gvd e : \estrt{r} \sreduces{i} \\ \lang{r'} \subseteq \lang{r}}
-%
-%\inferline{T-Check-NotOk}
-%{\gvd \echeck{r'}{e} : \estrt{r'} \sreduces \icheck{r}{i}}
-%{\gvd e : \estrt{r} \sreduces i \\ \lang{r'} \not \subseteq \lang{r}}
-%  
-%%\inferline{E-dconvert (optional) }
-%%{\dconvert{\strin[r']}{e} : \strin[r'] \ireduces \ifilter{r'}{\istrf{e}}}
-%%{e \val }
-%
-%\end{tabular}
-%\caption{Type-directed translation to internal terms for $\lcs$}
-%\label{fig:eeval}
-%\end{center}
-%\end{figure}
-
-%
-% Begin typing relation for internal terms.
-%
 \begin{figure}
@@
 \begin{tabular}{c r}
-\inferline{Ti-Var}
-{\ovd x : \sigma}
-{x : \sigma \in \Omega}
+\inferline{T-E-Str}
+{\tstr{s} \treduces \str{s}}
+{ \cdot }
 
-% Add in pedantic rules as well? If there's space.
-\inferline{Ti-Abs}
-{\ovd \lambda x : \sigma . i : \sigma \rightarrow \sigma'}
-{\Omega, x : \sigma \vdash i : \sigma'}
+\inferline{T-E-Include}
+{\rx{r} \treduces \rx{r}}
+{ \cdot }
 
-\inferline{Ti-App}
-{i_1 i_2 : \sigma'}
-{\ovd i_1 : \sigma \rightarrow \sigma' \\ \ovd i_2 : \sigma}
+\inferline{T-E-Concat}
+{\tconcat{T_1}{T_2} \treduces \tstr{s_1 \cdot s_2}} % ???
+{T_1 \treduces \tstr{s_1} \\ T_2 \treduces \tstr{s_2}}
 
-\inferline{Ti-Str}
-{\ovd \istr{s} : \istrt }
-{\cdot}
+\inferline{T-E-Subst}
+{\tsubst{r}{T_2}{T_3} \treduces \tstr{s}}
+{T_1 \treduces \rx{r} \\ T_2 \treduces \tstr{s_2} \\ T_3 \treduces \tstr{s_3} \\ \dosubst{r}{s_2}{s_3} = s}
 
-\inferline{Ti-Replace}
-{\ovd \ireplace{r}{i_1}{i_2} : \istrt}
-{\ovd i_1 : \istrt \\ \ovd i_2 : \istrt}
+\inferline{T-E-Check-OK}
+{\tcheck{r}{T} \treduces \tstr{s}}
+{T \treduces \strin{s} \\ s \in \lang{r}}
 
-\inferline{Ti-Check}
-{\ovd \icheck{r}{i_1}{i_2} : \istrt}
-{\ovd i_1 : \istrt \\ \ovd i_2 : \istrt}
-
-\inferline{Ti-Concat}
-{\ovd i_1 \iconcat i_2 : \istrt}
-{\ovd i_1 : \istrt \\ \ovd i_2 : \istrt}
+\inferline{T-E-Check-NotOk}
+{T \err}
+{\tcheck{r}{T} \treduces \tstr{s} \\ s \not \in \lang{r}}
 
 \end{tabular}
-\caption{Typing relation on internal terms.}\label{fig:itypes}
+\caption{Big step semantics for the target lanugage.
+There are also unimplemented error propagation rules for for Te-Concat and Te-Subst.}
+\label{fig:ieval}
 \end{center}
@@
 %
-\newcommand{\err}{{\tt err}}
+
 \begin{figure}
@@
 
-  \inferline{Ei-strval}
-  {\istr{s} \ival}
-  { \cdot }
+\inferline{Tr-string}
+{\trden{\strin{s}} = \tstr{s}}
+{ \ }
 
-  \inferline{Ei-Replace}
-  {\ireplace{r}{i_1}{i_2} \ireduces i'}
-  { i \ireduces \replacefn{r}{i_1}{i_2} }
+\inferline{Tr-Concat}
+{\trden{\siconcat{S_1}{S_2}} = \tconcat{T_1}{T_2}}
+{ \trden{S_1} = T_1 \\ \trden{S_2} = T_2}
 
-  \inferline{Ei-CheckOk}
-  {\icheck{r}{i} \ireduces i}
-  {i \in \lang{r}}
+\inferline{Tr-Subst}
+{\trden{ \sisubst{r}{S_1}{S_2} } = \tsubst{\rx{r}}{T_1}{T_2} }
+{ \trden{S_1} = T_1 \\ \trden{S_2} = T_2 }
 
-  \inferline{Ei-Check-NotOk}
-  {\icheck{r}{i} \ireduces \err}
-  {i \not \in \lang{r}}
+\inferline{Tr-Subst}
+{\trden{ \coerce{r'}{S} } = T}
+{ \trden{s} = T \\ S : \strin{r} \\ \lang{r'} \subseteq \lang{r}}
 
-  \inferline{Ei-Concat}
-  { i_1 \iconcat i_2 \ireduces \istr{i_1 \iconcat i_2} }
-  { \cdot }
+\inferline{Tr-Coerce-NotOk}
+{\trden{\coerce{r'}{S}} = \tcheck{\rx{r'}}{S}}
+{\trden{S} = T \\ s : \strin{r} \\ \lang{r'} \not \subseteq \lang{r}}
 
 \end{tabular}
-\caption{Big-step semantics for internal language; standard rules for the STL are omitted for brevity. The meta-variable $s$ ranges over string literals.}\label{fig:ieval}
+\caption{Translation from source terms (S) to target terms (T).
+The translation is type-directed in the Tr-Coerce cases.}
 \end{center}
@@
 Since sanitation problems are generally expressible with smalle, simple expressions,
-we believe that the compile-time overhead is not significant enough to prohibit
-our target usecase. Informal experimentation with our implementation seems to
+we believe that the compile-time overhead is not significant enough to prohibiour target usecase. Informal experimentation with our implementation seems to
 support this assertion.
@@
 \begin{thm}[Main Soundness Result]
-  If $\cdot \vdash e : \estrt{r} \sreduces i$ then either $i \ireduces \err$ or
+  If $\cdot \vdash e : \strin{r} \sreduces i$ then either $i \treduces \err$ or
   if $i \sreduces v$ for $v \ival$ then $v \in r$.
@@
 %  \begin{itemize}
-%    \item If $e:T$ and $e:T \ireduces i$ then $i : \sigma$ such that $h(T) = \sigma$.
+%    \item If $e:T$ and $e:T \treduces i$ then $i : \sigma$ such that $h(T) = \sigma$.
 %    \item If $e:T$ and $e:T \reduces e'$ then $e':T$.
@@
 %  \item \textbf{E-Stringval, E-strval}. Both cases hold since no reduction is possible. 
-%  \item \textbf{E-String}. By the definition of typing for internal terms, $\istrf{e} : istr$. It suffices to show that $h(\str) = istr$, which follows from the definition of $h$.
+%  \item \textbf{E-String}. By the definition of typing for internal terms, $\tstrf{e} : istr$. It suffices to show that $h(\str) = istr$, which follows from the definition of $h$.
 %  \item \textbf{E-String\_in}.
-By the definition of typing for internal terms, $\istrf{e} : istr$.
+By the definition of typing for internal terms, $\tstrf{e} : istr$.
 It suffices to show that $h(\strin[r]) = istr$, which follows from the definition of $h$ for arbitrary $r$.
 %  \item \textbf{E-concatval}.
-By the definition of typing for internal terms, $\istrf{e_1} + \istrf{e_2} : istr$.
+By the definition of typing for internal terms, $\tstrf{e_1} + \tstrf{e_2} : istr$.
 So it suffices to show that $h(\strin[r_1+r_2]) = istr$, which follows from the definition of $h$ for arbitrary $r_1$, $r_2$.
@@
 %  \item \textbf{E-DConvertval}.
-By the definition of typing for internal terms, $\ifilter(r,\istrf{e}) : istr$.
+By the definition of typing for internal terms, $\ifilter(r,\tstrf{e}) : istr$.
 It suffices to show that $h(\strin[r_1+r_2]) = istr$, which follows from the definition of $h$.
@@
 %\begin{thm}[Progress]
-%  If $e:T$ then $e:T \reduces^* e' \ireduces^* i$ where $i \ival$.
+%  If $e:T$ then $e:T \reduces^* e' \treduces^* i$ where $i \ival$.
 %\end{thm}
@@
 %For \textbf{T-Equiv-Include}, note that $e \in \lang{r}$, $e = \strf{s}$ for some $s$; therefore, $e \val$ by E-Strinval. 
-%We have that $e \val$ and $e : \strin[r]$, so $e \ireduces \istrf{e}$ by E-string\_in.
+%We have that $e \val$ and $e : \strin[r]$, so $e \treduces \tstrf{e}$ by E-string\_in.
 %The remaining cases follow by induction on the hypotheses and application of corresponding evaluation rules.
@@
 \hline
-$\estrt{r}$ & & $\fvar{stringin}[r]$ & \verb|string_in["|$r$\verb|"]|\\
+$\strin{r}$ & & $\fvar{stringin}[r]$ & \verb|string_in["|$r$\verb|"]|\\
 \\
@@
 \end{document}
-
-\end{document}

### 2014-04-16 sanitation-plas14/sanitation-plas14.tex
P -> T, but macros still say T :-(

@@
 \newcommand{\err}{{\tt err}}
-\newcommand{\trden}[1]{[[#1]]} % = Translation Denotation.
+\newcommand{\trden}[1]{\llbracket #1 \rrbracket} % = Translation Denotation.
 
@@
 
-<T> ::= \hfill Internal expressions \\
-  $\tstr{T}$ \alt
-  $\siconcat{T}{T}$ \alt
-  $\tsubst{r}{T}{T}$ \alt
-  $\tcheck{r}{T}$
+<P> ::= \hfill Internal expressions \\
+  $\tstr{P}$ \alt
+  $\siconcat{P}{P}$ \alt
+  $\tsubst{r}{P}{P}$ \alt
+  $\tcheck{r}{P}$
 \end{grammar}
@@
 
-\inferline{T-T-String}
+\inferline{P-T-String}
 {\tctx \vdash \tstr{s} : \str}
@@
 
-\inferline{T-T-Concat}
-{\tctx \vdash \tconcat{T_1}{T_2} : \str}
-{\tctx \vdash T_1 : \str \\ \tctx \vdash T_2 : \str}
+\inferline{P-T-Concat}
+{\tctx \vdash \tconcat{P_1}{P_2} : \str}
+{\tctx \vdash P_1 : \str \\ \tctx \vdash P_2 : \str}
 
-\inferline{T-T-Subst}
-{\tctx \vdash \tsubst{\rx{r}}{T_1}{T_2} : \str}
-{\tctx \vdash T_1 : \str \\ \tctx \vdash T_2 : \str}
+\inferline{P-T-Subst}
+{\tctx \vdash \tsubst{\rx{r}}{P_1}{P_2} : \str}
+{\tctx \vdash P_1 : \str \\ \tctx \vdash P_2 : \str}
 
-\inferline{T-T-Check}
+\inferline{P-T-Check}
 {\tctx \vdash \tcheck{\rx{r}}{s} : \str}
-{\tctx \vdash T : \str}
+{\tctx \vdash P : \str}
 
 \end{tabular}
-\caption{Typing rules for target langauge terms (T). The metavariable s ranges over
+\caption{Typing rules for target langauge terms (P). The metavariable s ranges over
 string literals.}
@@
 \begin{tabular}{c r}
-\inferline{T-E-Str}
+\inferline{P-E-Str}
 {\tstr{s} \treduces \str{s}}
@@
 
-\inferline{T-E-Include}
+\inferline{P-E-Include}
 {\rx{r} \treduces \rx{r}}
@@
 
-\inferline{T-E-Concat}
-{\tconcat{T_1}{T_2} \treduces \tstr{s_1 \cdot s_2}} % ???
-{T_1 \treduces \tstr{s_1} \\ T_2 \treduces \tstr{s_2}}
+\inferline{P-E-Concat}
+{\tconcat{P_1}{P_2} \treduces \tstr{s_1 \cdot s_2}} % ???
+{P_1 \treduces \tstr{s_1} \\ P_2 \treduces \tstr{s_2}}
 
-\inferline{T-E-Subst}
-{\tsubst{r}{T_2}{T_3} \treduces \tstr{s}}
-{T_1 \treduces \rx{r} \\ T_2 \treduces \tstr{s_2} \\ T_3 \treduces \tstr{s_3} \\ \dosubst{r}{s_2}{s_3} = s}
+\inferline{P-E-Subst}
+{\tsubst{r}{P_2}{P_3} \treduces \tstr{s}}
+{P_1 \treduces \rx{r} \\ P_2 \treduces \tstr{s_2} \\ P_3 \treduces \tstr{s_3} \\ \dosubst{r}{s_2}{s_3} = s}
 
-\inferline{T-E-Check-OK}
-{\tcheck{r}{T} \treduces \tstr{s}}
-{T \treduces \strin{s} \\ s \in \lang{r}}
+\inferline{P-E-Check-OK}
+{\tcheck{r}{P} \treduces \tstr{s}}
+{P \treduces \strin{s} \\ s \in \lang{r}}
 
-\inferline{T-E-Check-NotOk}
-{T \err}
+\inferline{P-E-Check-NotOk}
+{P \err}
 {\tcheck{r}{T} \treduces \tstr{s} \\ s \not \in \lang{r}}
@@
 \inferline{Tr-Concat}
-{\trden{\siconcat{S_1}{S_2}} = \tconcat{T_1}{T_2}}
-{ \trden{S_1} = T_1 \\ \trden{S_2} = T_2}
+{\trden{\siconcat{S_1}{S_2}} = \tconcat{P_1}{P_2}}
+{ \trden{S_1} = P_1 \\ \trden{S_2} = P_2}
 
 \inferline{Tr-Subst}
-{\trden{ \sisubst{r}{S_1}{S_2} } = \tsubst{\rx{r}}{T_1}{T_2} }
-{ \trden{S_1} = T_1 \\ \trden{S_2} = T_2 }
+{\trden{ \sisubst{r}{S_1}{S_2} } = \tsubst{\rx{r}}{P_1}{P_2} }
+{ \trden{S_1} = P_1 \\ \trden{S_2} = P_2 }
 
 \inferline{Tr-Subst}
-{\trden{ \coerce{r'}{S} } = T}
-{ \trden{s} = T \\ S : \strin{r} \\ \lang{r'} \subseteq \lang{r}}
+{\trden{ \coerce{r'}{S} } = P}
+{ \trden{s} = P \\ S : \strin{r} \\ \lang{r'} \subseteq \lang{r}}
 
@@
 {\trden{\coerce{r'}{S}} = \tcheck{\rx{r'}}{S}}
-{\trden{S} = T \\ s : \strin{r} \\ \lang{r'} \not \subseteq \lang{r}}
+{\trden{S} = P \\ s : \strin{r} \\ \lang{r'} \not \subseteq \lang{r}}
 
 \end{tabular}
-\caption{Translation from source terms (S) to target terms (T).
+\caption{Translation from source terms (S) to target terms (P).
 The translation is type-directed in the Tr-Coerce cases.}
@@
 that a check must succeed. Furthermore, expensive calculations (such as language
-inclusion or ${\tt computer\_replacement\_language}$, all occur at compile time.
+inclusion or ${\tt dsubst}$, all occur at compile time.
 Since sanitation problems are generally expressible with smalle, simple expressions,

### 2014-04-16 sanitation-plas14/sanitation-plas14.tex
Removed val judgements.

@@
 
-\newcommand{\val}{ \ {\tt val} }
-\newcommand{\ival}{ \ {\tt ival} }
-
 \newcommand{\tstrf}[1]{`#1\textrm'} %??
@@
 The semantics are defined in terms of an internal language with at least strings and a regex filter function.
-These constraints are captured by the internal term valuations ($\ival$).
+These constraints are captured by the internal term valuations.
 The internal language does not necessarily need a regex filter function because
@@
 \begin{tabular}{c r}
-\inferline{Te-Include}
-{\strin{s} \val}
+\inferline{S-E-Include}
+{\strin{s} \sreduces \strin{s}}
 { \cdot }
@@
 \inferline{S-E-Concat}
-{\siconcat{S_1}{S_2} \ereduces \strin{s_1 \cdot s_2}} % ???
-{S_1 \Downarrow \strin{s_1} \\ S_2 \Downarrow \strin{s_2}}
+{\siconcat{S_1}{S_2} \sreduces \strin{s_1 \cdot s_2}} % ???
+{S_1 \sreduces \strin{s_1} \\ S_2 \sreduces \strin{s_2}}
 
 \inferline{S-E-Subst}
-{\sisubst{r}{S_1}{S_2} \ereduces s}
-{S_1 \Downarrow \strin{s_1} \\ S_2 \Downarrow \strin{s_2} \\ \dosubst{r}{s_1}{s_2} = s}
+{\sisubst{r}{S_1}{S_2} \sreduces s}
+{S_1 \sreduces \strin{s_1} \\ S_2 \sreduces \strin{s_2} \\ \dosubst{r}{s_1}{s_2} = s}
 
 \inferline{S-E-Coerce-OK}
-{\coerce{r}{S} \ereduces \strin{s}}
-{S \ereduces \strin{s} \\ x \in \lang{r}}
+{\coerce{r}{S} \sreduces \strin{s}}
+{S \sreduces \strin{s} \\ x \in \lang{r}}
 
@@
 {S \err}
-{\coerce{r}{S} \ereduces \strin{s} \\ s \not \in \lang{r}}
+{\coerce{r}{S} \sreduces \strin{s} \\ s \not \in \lang{r}}
 
@@
   If $\cdot \vdash e : \strin{r} \sreduces i$ then either $i \treduces \err$ or
-  if $i \sreduces v$ for $v \ival$ then $v \in r$.
+  if $i \sreduces v$ for $v$??? then $v \in r$.
 \end{thm}

### 2014-04-16 sanitation-plas14/sanitation-plas14.tex
typo.
Fixed errors and typos in the rules.
Using new style for rules.

@@
 % Judgements
-\newcommand{\err}{{\tt err}}
+\newcommand{\err}{\ {\tt err}}
 \newcommand{\trden}[1]{\llbracket #1 \rrbracket} % = Translation Denotation.
@@
             Carnegie Mellon University}
-           {\{nfulton, comar, aldrich\}@cs.cmu.edu}
+           {\{nathanfu, comar, aldrich\}@cs.cmu.edu}
 
@@
 casting.
-
-% Whatever just dropping this in somewhere random.
-\begin{thm}
-If $S : \strin{r}$ then there exists a T such that $\trden{s} = T$ and either: 
-\begin{itemize}
-\item (a) $T \treduces \tstr{s}$ and $S \sreduces \strin{s}$, and $s \in lang{r}$.
-\item (b) $T \err$ and $S \err$.
-\end{itemize}
-\end{thm}
-\begin{proof}
-By induction on the derivation of $\trden{S} = T$. TODO-nrf add case magic.
-\begin{itemize}
-\item Tr-string. By assumption $\trden{S} = \tstr{s}$. The (a) property follows.
-By S-E-Str, $S \sreduces \str{s}$ and by T-E-Str, $\tstr{s} \treduces \tstr{s}$.
-By source language type safety, it follows that $s \in \lang{r}$.
-\item Tr-Concat.
-By inversion of the typing relaiton for the source lnaguage, $S_1 : \strin{r_1}$ and $S_2 : \stringin{S_2}$.
-By induction on the left premise, $T_1 \treduces \tstr{s_1}$, $S_1 \sreduces \str{s_1}$ and $s_1 \in \lang{r_1}$
-or else $T_1 err$. 
-Similarly for $T_2$ and $S_2$.
-In the error cases, the propagations rules corresponding to S-E-Concat and T-E-Concat imply that $\tconcat{T_1}{T_2} \err$ and $\siconcat{S_1}{S_2} \err$.
-In the non-error case, $\siconcat{S_1}{S_2} \sreduces \strin{s_1 \cdot s_2}$ by S-E-Concat and $\tconcat{T_1}{T_2} \treduces \tstr{s_1 \cdot s_2}$ by T-E-Concat.
-Finally, $s_1 \in \lang{r_1}$ and $s_2 \in \lang{r_2}$ implies that $s_1 \cdot s_2 \in \lang{r}$ by \ref{lemmaregex}.
-\item T-subst.
-Suppose $\sisubst{r}{S_1}{S_2} : \strin{s}$ for some $x$. 
-By inverison, $S_1 : \strin{r_1}$ and $S_2 : \strin{r_2}$.
-Note that $s : \subst{r}{S_1}{S_2} : \stringin{\tsubst{r}{r_1}{r_2}}$ by S-T-Subst.
-By induction, $\trden{S_1} = T_1$ and $\trden{S_2} = T_2$.
-Either (a) or (b) holds for each.
-
-If (a) holds for both, the $S_1 \sreduces \strin{S_1}$ and $S_2 \sreduces \strin{S_2}$ for $s_1 \in \lang{r_1}$ and $s_2 \in \lang{r_2}$.
-Therefore, $\sisubst{r}{S_1}{S_2} \sreduces \strin{\dosubst{r,s_1,s_2}}$.
-The same fact can be shown in an analogous manner for $T_1$ and $T_2$.
-Finally, $s_1 \in \lang{r_1}$ and $s_2 \in \lang{r_2}$, which implies that $\dosubst{r}{s_1}{s_2} \in \lang{\sisubst{r}{s_1}{s_2}}$ by \ref{substhm}.
-
-If (b) holds for either the left or right premise, then $\sisubst{r}{S_1}{S_2} \err$ by the propagation rules corresponding to S-T-subst, 
-and $\tsubst{S_1}{S_2} \err$ by the propagation rules corresponding to T-T-Subst.
-\end{itemize}
-\end{proof}
-
 %
@@
 %
-\subsection{Properties of Regular Languages}
-
-The regular languages are the smallest set generated by regular expressions
-defined in Figure 1.
-
-\begin{thm}{Closure Properties.} \label{thm:closure}
-The regular expressions are closed under complements and concatenation.
-\end{thm}
-\begin{proof}
-See \cite{cinderella}.
-\end{proof}
-
-\begin{thm}{Coercion Theorem.} \label{thm:coerce}
-Suppose that $R$ and $L$ are regular expressions, and that $s \in R$ is a finite string.
-Let $s' := coerce(R,L,s)$ with all maximal substrings recognized by $L$ replaced with $\epsilon$.
-Then $s'$ is recognized by $(R \backslash L) + \emptyset$ and the construction of $R \backslash L$ is decidable.
-\end{thm}
-\begin{proof}
-Let $F,G$ be FAs corresponding to $R$ and $L$, and let $G'$ be $G$ with its final states inverted (so that $G'$ is the complement of $L$).
-Define an FA $H$ as a DFA corresponding to the NFA found by combining $F$ and $G'$ such that $H$ accepts only if $R$ and $L'$ accept or if $s$ is empty (this construction may result in an exponential blowup in state size.)  Clearly, $H$ corresponds to $R \backslash L + \emptyset$.
-Thus, the construction of $R \backslash L + \emptyset$ is decidable.
-
-If $R \subset L$, $s' = \emptyset$.
-If $L \subset R$, either $s' = \emptyset$, or $s' \in R$ and $s' \not \in L$.
-If $R$ and $L$ are not subsets of one another, then it may be the case that $L$ recognizes part of $R$.
-Consider $L$ as the union of two languages, one which is a subset of $R$ and one which is disjoint.
-The subset language is considered above and the disjoint language is inconsequential.
-\end{proof}
 
@@
 
-\onecolumn
 \newcommand{\sctx}{\Gamma} % Context for external typing
@@
 
-\begin{figure}
-\begin{center}
-\begin{tabular}{c r}
+\begin{figure}[t]
+\small
+$\fbox{\inferrule{}{\sctx \vdash S : \psi}}$
+~~~~$\sctx ::= \emptyset \pipe \sctx, x : \psi$
+\begin{mathpar}
 
-\inferline{S-T-include}
+\inferrule[S-T-include]
+{s \in \lang{r}}
 {\sctx \vdash \strin{s} : \stringin{r}}
-{s \in \lang{r}}
 
-\inferline{S-T-Concat}
-{\sctx \vdash \siconcat{S_1}{S_2} : \strin{r_1 \cdot r_2}}
-{\sctx \vdash S_1 : \strin{r_1} \\ \sctx \vdash S_2 : \strin{r_2}}
+\inferrule[S-T-Concat]
+{\sctx \vdash S_1 : \stringin{r_1} \\ \sctx \vdash S_2 : \stringin{r_2}}
+{\sctx \vdash \siconcat{S_1}{S_2} : \stringin{r_1 \cdot r_2}}
 
-\inferline{S-T-subst}
-{\sctx \vdash \sisubst{r}{S_1}{S_2} : r'}
-{\sctx \vdash S_1 : \strin{r_1} \\ \sctx \vdash S_2 : \strin{r_2} \\ {\tt lsubst}(r,r_1,r_2) = r'}
+\inferrule[S-T-subst]
+{\sctx \vdash S_1 : \stringin{r_1} \\ \sctx \vdash S_2 : \stringin{r_2} \\ {\tt lsubst}(r,r_1,r_2) = r'}
+{\sctx \vdash \sisubst{r}{S_1}{S_2} : \stringin{r'}}
 
-\inferline{S-T-coerce}
+\inferrule[S-T-coerce]
+{\sctx \vdash S : \stringin{r}}
 {\sctx \vdash \coerce{r'}{s} : \stringin{r'}}
-{\sctx \vdash S : \stringin{r}}
 
-\end{tabular}
+\end{mathpar}
+
 \caption{Typing rules for source langauge terms (S). The metavariable s ranges over
@@
 \label{fig:etyping}
-\end{center}
 \end{figure}
 
-\begin{figure}
-\begin{center}
-\begin{tabular}{c r}
-\inferline{S-E-Include}
+\begin{figure}[t]
+\small
+$\fbox{\inferrule{}{S \sreduces S}}$
+\begin{mathpar}
+
+\inferrule[S-E-Include]
+{ \ }
 {\strin{s} \sreduces \strin{s}}
-{ \cdot }
 
-\inferline{S-E-Concat}
+\inferrule[S-E-Concat]
+{S_1 \sreduces \strin{s_1} \\ S_2 \sreduces \strin{s_2}} 
 {\siconcat{S_1}{S_2} \sreduces \strin{s_1 \cdot s_2}} % ???
-{S_1 \sreduces \strin{s_1} \\ S_2 \sreduces \strin{s_2}}
 
-\inferline{S-E-Subst}
+\inferrule[S-E-Subst]
+{S_1 \sreduces \strin{s_1} \\ S_2 \sreduces \strin{s_2} \\ \dosubst{r}{s_1}{s_2} = s} 
 {\sisubst{r}{S_1}{S_2} \sreduces s}
-{S_1 \sreduces \strin{s_1} \\ S_2 \sreduces \strin{s_2} \\ \dosubst{r}{s_1}{s_2} = s}
 
-\inferline{S-E-Coerce-OK}
+\inferrule[S-E-Coerce-OK]
+{S \sreduces \strin{s} \\ s \in \lang{r}}
 {\coerce{r}{S} \sreduces \strin{s}}
-{S \sreduces \strin{s} \\ x \in \lang{r}}
 
-\inferline{S-E-Coerce-NotOk}
+\inferrule[S-E-Coerce-NotOk]
+{\coerce{r}{S} \sreduces \strin{s} \\ s \not \in \lang{r}}
 {S \err}
-{\coerce{r}{S} \sreduces \strin{s} \\ s \not \in \lang{r}}
 
-\end{tabular}
+\end{mathpar}
 \caption{Big step semantics for the string fragment of the source language.
@@
 \label{fig:eeval}
-\end{center}
 \end{figure}
 
-\begin{figure}
-\begin{center}
-\begin{tabular}{c r}
+\newpage
 
-\inferline{P-T-String}
+\begin{figure}[t]
+\small
+$\fbox{\inferrule{}{\tctx \vdash P : \theta}}$
+~~~~$\tctx ::= \emptyset \pipe \tctx, x : \theta$
+
+\begin{mathpar}
+
+\inferrule[P-T-String]
+{ \ }
 {\tctx \vdash \tstr{s} : \str}
-{\cdot}
 
-\inferline{P-T-Concat}
+\inferrule[P-T-Concat]
+{\tctx \vdash P_1 : \str \\ \tctx \vdash P_2 : \str}
 {\tctx \vdash \tconcat{P_1}{P_2} : \str}
+
+\inferrule[P-T-Subst]
 {\tctx \vdash P_1 : \str \\ \tctx \vdash P_2 : \str}
+{\tctx \vdash \tsubst{\rx{r}}{P_1}{P_2} : \str}
 
-\inferline{P-T-Subst}
-{\tctx \vdash \tsubst{\rx{r}}{P_1}{P_2} : \str}
-{\tctx \vdash P_1 : \str \\ \tctx \vdash P_2 : \str}
+\inferrule[P-T-Check]
+{\tctx \vdash P : \str}
+{\tctx \vdash \tcheck{\rx{r}}{s} : \str}
 
-\inferline{P-T-Check}
-{\tctx \vdash \tcheck{\rx{r}}{s} : \str}
-{\tctx \vdash P : \str}
-
-\end{tabular}
+\end{mathpar}
 \caption{Typing rules for target langauge terms (P). The metavariable s ranges over
@@
 \label{fig:etyping}
-\end{center}
 \end{figure}
 
-\begin{figure}
-\begin{center}
-\begin{tabular}{c r}
-\inferline{P-E-Str}
-{\tstr{s} \treduces \str{s}}
-{ \cdot }
+\begin{figure}[t]
+\small
+$\fbox{\inferrule{}{P \treduces P}}$
+\begin{mathpar}
+\inferrule[P-E-Strin]
+{ \ }
+{\rx{r} \treduces \rx{r}}
 
-\inferline{P-E-Include}
-{\rx{r} \treduces \rx{r}}
-{ \cdot }
+\\
 
-\inferline{P-E-Concat}
+\inferrule[P-E-Concat]
+{P_1 \treduces \tstr{s_1} \\ P_2 \treduces \tstr{s_2}} 
 {\tconcat{P_1}{P_2} \treduces \tstr{s_1 \cdot s_2}} % ???
-{P_1 \treduces \tstr{s_1} \\ P_2 \treduces \tstr{s_2}}
 
-\inferline{P-E-Subst}
+\inferrule[P-E-Subst]
+{P_1 \treduces \rx{r} \\ P_2 \treduces \tstr{s_2} \\ P_3 \treduces \tstr{s_3} \\ \dosubst{r}{s_2}{s_3} = s} 
 {\tsubst{r}{P_2}{P_3} \treduces \tstr{s}}
-{P_1 \treduces \rx{r} \\ P_2 \treduces \tstr{s_2} \\ P_3 \treduces \tstr{s_3} \\ \dosubst{r}{s_2}{s_3} = s}
 
-\inferline{P-E-Check-OK}
+\inferrule[P-E-Check-OK]
+{P \treduces \strin{s} \\ s \in \lang{r}}
 {\tcheck{r}{P} \treduces \tstr{s}}
-{P \treduces \strin{s} \\ s \in \lang{r}}
 
-\inferline{P-E-Check-NotOk}
+\inferrule[P-E-Check-NotOk]
+{\tcheck{r}{P} \treduces \tstr{s} \\ s \not \in \lang{r}}
 {P \err}
-{\tcheck{r}{T} \treduces \tstr{s} \\ s \not \in \lang{r}}
-
-\end{tabular}
+\end{mathpar}
 \caption{Big step semantics for the target lanugage.
-There are also unimplemented error propagation rules for for Te-Concat and Te-Subst.}
+There are also unimplemented error propagation rules for for P-E-Concat and P-E-Subst.}
 \label{fig:ieval}
-\end{center}
 \end{figure}
 
+\newpage
+
 %
-% Begin big-step sematnics for internal terms.
+% Begin translation
 %
 
-\begin{figure}
-\begin{center}
-\begin{tabular}{c r}
+\begin{figure}[t]
 
-\inferline{Tr-string}
+$\fbox{\inferrule{}{\trden{S} = P}}$
+\begin{mathpar}
+
+\inferrule[Tr-string]
+{ \ }
 {\trden{\strin{s}} = \tstr{s}}
-{ \ }
 
-\inferline{Tr-Concat}
+\inferrule[Tr-Concat]
+{ \trden{S_1} = P_1 \\ \trden{S_2} = P_2}
 {\trden{\siconcat{S_1}{S_2}} = \tconcat{P_1}{P_2}}
-{ \trden{S_1} = P_1 \\ \trden{S_2} = P_2}
 
-\inferline{Tr-Subst}
+\inferrule[Tr-Subst]
+{ \trden{S_1} = P_1 \\ \trden{S_2} = P_2 }
 {\trden{ \sisubst{r}{S_1}{S_2} } = \tsubst{\rx{r}}{P_1}{P_2} }
-{ \trden{S_1} = P_1 \\ \trden{S_2} = P_2 }
 
-\inferline{Tr-Subst}
-{\trden{ \coerce{r'}{S} } = P}
-{ \trden{s} = P \\ S : \strin{r} \\ \lang{r'} \subseteq \lang{r}}
+\inferrule[Tr-Coerce-Ok]
+{ S : \strin{r} \\ \lang{r'} \subseteq \lang{r}}
+{\trden{ \coerce{r'}{S} } = \tstr{s}}
 
-\inferline{Tr-Coerce-NotOk}
-{\trden{\coerce{r'}{S}} = \tcheck{\rx{r'}}{S}}
-{\trden{S} = P \\ s : \strin{r} \\ \lang{r'} \not \subseteq \lang{r}}
+\inferrule[Tr-Coerce-NotOk]
+{\trden{S} = P \\ S : \strin{r} \\ \lang{r'} \not \subseteq \lang{r}}
+{\trden{\coerce{r'}{S}} = \tcheck{\rx{r'}}{P}}
 
-\end{tabular}
+\end{mathpar}
 \caption{Translation from source terms (S) to target terms (P).
 The translation is type-directed in the Tr-Coerce cases.}
-\end{center}
 \end{figure}
-
-\twocolumn
 
@@
 generally well-tested and there exist verified implementations.
+
+\subsection{Properties of Regular Languages}
+
+The soundness theorems in this section depend upon some well-known properties
+of regular expressions. Throughout this section, suppose regular expression
+$r$ and strings $s$ are defined over a single, fixed, finite alphabet $\Sigma$.
+
+\begin{thm} \label{thm:closure}
+The regular expressions are closed under complements and concatenation.
+\end{thm}
+\begin{proof}
+See \cite{cinderella}.
+\end{proof}
+
+\begin{defn}[$\tt{subst}$]
+  The function $\subst{r}{s_1}{s_2}$ produces a string in which all substrings
+  of $s_1$ matching $r$ are replaced with $s_2$.
+\end{defn}
+
+\begin{defn}[$\tt{lsubst}$]
+  The function $\lsubst{r}{r_1}{r_2}$ produces a regular expression in which
+  and sublanguage $r_1'$ of $r_1$ satisfying the condition $\lang{r_1'} \subseteq{r}$
+  is replaced with $r_2$.
+\end{defn}
+
+We now present some properties about substitution.
+
+\begin{thm}
+  String substitution ($\subst{r}{s_1}{s_2}$) is decidable and defined for all expressions
+  and strings over similar alpha.
+\end{thm}
+
+\begin{thm}
+  Language substittion $(\lsubst{r}{r_1}{r_2})$ is decidable and defined for
+  all expressions.
+\end{thm}
+
+The following relationship between string substituion and language substitution is at
+the heart of our main correctness reesult.
+
+\begin{thm}
+  For all $r$, if $s_1 \in r_1$ and $s_2 \in r_2$ then 
+  $\subst{r}{s_1}{s_2} \subseteq \lsubst{r}{r_1}{r_2}$.
+\end{thm}
+
+\begin{lem}{Coercion Lemma.} \label{thm:coercionalemma}
+Suppose that $R$ and $L$ are regular expressions, and that $s \in R$ is a finite string.
+Let $s' := coerce(R,L,s)$ with all maximal substrings recognized by $L$ replaced with $\epsilon$.
+Then $s'$ is recognized by $(R \backslash L) + \emptyset$ and the construction of $R \backslash L$ is decidable.
+\end{lem}
+\begin{proof}
+Let $F,G$ be FAs corresponding to $R$ and $L$, and let $G'$ be $G$ with its final states inverted (so that $G'$ is the complement of $L$).
+Define an FA $H$ as a DFA corresponding to the NFA found by combining $F$ and $G'$ such that $H$ accepts only if $R$ and $L'$ accept or if $s$ is empty (this construction may result in an exponential blowup in state size.)  Clearly, $H$ corresponds to $R \backslash L + \emptyset$.
+Thus, the construction of $R \backslash L + \emptyset$ is decidable.
+
+If $R \subset L$, $s' = \emptyset$.
+If $L \subset R$, either $s' = \emptyset$, or $s' \in R$ and $s' \not \in L$.
+If $R$ and $L$ are not subsets of one another, then it may be the case that $L$ recognizes part of $R$.
+Consider $L$ as the union of two languages, one which is a subset of $R$ and one which is disjoint.
+The subset language is considered above and the disjoint language is inconsequential.
+\end{proof}
+
+\subsection{Safety of the Source and Target Languages}
+
+We now prove safety for the string fragment of the source and target languages.
+
+\begin{thm}
+  Let $S$ be a term in the source language. If $\sctx \vdash S : \psi$ 
+  then $S \sreduces S'$ and $S' : \psi$.
+\end{thm}
+
+\begin{thm}
+  Let $P$ be a term in the target language. If $\sctx \vdash P : \theta$ 
+  then $P \sreduces P'$ and $P' : \theta$.
+\end{thm}
+
+\subsection{Translation Correctness}
+
+We now present the main correctness result.
+
+\begin{thm}
+If $S : \strin{r}$ then there exists a T such that $\trden{s} = T$ and either: 
+\begin{itemize}
+\item (a) $T \treduces \tstr{s}$ and $S \sreduces \strin{s}$, and $s \in lang{r}$.
+\item (b) $T \err$ and $S \err$.
+\end{itemize}
+\end{thm}
+\begin{proof}
+By induction on the derivation of $\trden{S} = T$. TODO-nrf add case magic.
+\begin{itemize}
+\item Tr-string. By assumption $\trden{S} = \tstr{s}$. The (a) property follows.
+By S-E-Str, $S \sreduces \str{s}$ and by T-E-Str, $\tstr{s} \treduces \tstr{s}$.
+By source language type safety, it follows that $s \in \lang{r}$.
+\item Tr-Concat.
+By inversion of the typing relaiton for the source lnaguage, $S_1 : \strin{r_1}$ and $S_2 : \stringin{S_2}$.
+By induction on the left premise, $T_1 \treduces \tstr{s_1}$, $S_1 \sreduces \str{s_1}$ and $s_1 \in \lang{r_1}$
+or else $T_1 err$. 
+Similarly for $T_2$ and $S_2$.
+In the error cases, the propagations rules corresponding to S-E-Concat and T-E-Concat imply that $\tconcat{T_1}{T_2} \err$ and $\siconcat{S_1}{S_2} \err$.
+In the non-error case, $\siconcat{S_1}{S_2} \sreduces \strin{s_1 \cdot s_2}$ by S-E-Concat and $\tconcat{T_1}{T_2} \treduces \tstr{s_1 \cdot s_2}$ by T-E-Concat.
+Finally, $s_1 \in \lang{r_1}$ and $s_2 \in \lang{r_2}$ implies that $s_1 \cdot s_2 \in \lang{r}$ by \ref{lemmaregex}.
+\item T-subst.
+Suppose $\sisubst{r}{S_1}{S_2} : \strin{s}$ for some $x$. 
+By inverison, $S_1 : \strin{r_1}$ and $S_2 : \strin{r_2}$.
+Note that $s : \subst{r}{S_1}{S_2} : \stringin{\tsubst{r}{r_1}{r_2}}$ by S-T-Subst.
+By induction, $\trden{S_1} = T_1$ and $\trden{S_2} = T_2$.
+Either (a) or (b) holds for each.
+
+If (a) holds for both, the $S_1 \sreduces \strin{S_1}$ and $S_2 \sreduces \strin{S_2}$ for $s_1 \in \lang{r_1}$ and $s_2 \in \lang{r_2}$.
+Therefore, $\sisubst{r}{S_1}{S_2} \sreduces \strin{\dosubst{r,s_1,s_2}}$.
+The same fact can be shown in an analogous manner for $T_1$ and $T_2$.
+Finally, $s_1 \in \lang{r_1}$ and $s_2 \in \lang{r_2}$, which implies that $\dosubst{r}{s_1}{s_2} \in \lang{\sisubst{r}{s_1}{s_2}}$ by \ref{substhm}.
+
+If (b) holds for either the left or right premise, then $\sisubst{r}{S_1}{S_2} \err$ by the propagation rules corresponding to S-T-subst, 
+and $\tsubst{S_1}{S_2} \err$ by the propagation rules corresponding to T-T-Subst.
+\end{itemize}
+\end{proof}
 

### 2014-04-16 sanitation-plas14/sanitation-plas14.tex
PLAS version used in today's meeting with JA.

@@
 % authoryear    To obtain author/year citation style instead of numeric.
+\usepackage{afterpage}
 \usepackage{listings}
@@
     \makeatletter
+\usepackage{enumitem}
+\usepackage{enumerate}
 
@@
 
+\afterpage{
+
 \begin{figure}[t]
@@
 
+\clearpage
+} % End \afterpage
+
 \newpage
@@
 
+\begin{lem}
+  If $\sctx \vdash S : \stringin[r]$ then $r$ is a well-formed regular expression. 
+\end{lem}
+\begin{proof}
+  By induction on the typing derivation and the closure lemmas for regular
+  expressions.
+\end{proof}
+
 We now prove safety for the string fragment of the source and target languages.
@@
 \end{thm}
+\begin{proof}
+  TODO-nrf
+
+By a straightforward induction on the derivation of $\sctx \vdash S : \psi$.
+Note that preservation depends upon the substitution correspondence theorem
+for the T-T-Subst case and the closure lemmas for the T-T-Concat case.
+\end{proof}
 
 \begin{thm}
-  Let $P$ be a term in the target language. If $\sctx \vdash P : \theta$ 
+  Let $P$ be a term in the target language. If $\tctx \vdash P : \theta$ 
   then $P \sreduces P'$ and $P' : \theta$.
 \end{thm}
+\begin{proof}
+TODO-nrf
+
+  By a straightforward induction on the derivation of $\tctx \vdash P : \theta$.
+Note that the well-formedness of regular expressions indexing ${\tt rx}$ types
+depends upon the closure lemmas.
+\end{proof}
 
@@
 If $S : \strin{r}$ then there exists a T such that $\trden{s} = T$ and either: 
-\begin{itemize}
-\item (a) $T \treduces \tstr{s}$ and $S \sreduces \strin{s}$, and $s \in lang{r}$.
-\item (b) $T \err$ and $S \err$.
-\end{itemize}
+\begin{description}
+  \item[(a)] $T \treduces \tstr{s}$ and $S \sreduces \strin{s}$, and $s \in lang{r}$.
+  \item[(b)] $T \err$ and $S \err$.
+  \end{description}
 \end{thm}
@@
 By induction on the derivation of $\trden{S} = T$. TODO-nrf add case magic.
-\begin{itemize}
-\item Tr-string. By assumption $\trden{S} = \tstr{s}$. The (a) property follows.
+\begin{description}
+\item[Tr-string] By assumption $\trden{S} = \tstr{s}$. The (a) property follows.
 By S-E-Str, $S \sreduces \str{s}$ and by T-E-Str, $\tstr{s} \treduces \tstr{s}$.
 By source language type safety, it follows that $s \in \lang{r}$.
-\item Tr-Concat.
+\item[Tr-Concat]
 By inversion of the typing relaiton for the source lnaguage, $S_1 : \strin{r_1}$ and $S_2 : \stringin{S_2}$.
@@
 Finally, $s_1 \in \lang{r_1}$ and $s_2 \in \lang{r_2}$ implies that $s_1 \cdot s_2 \in \lang{r}$ by \ref{lemmaregex}.
-\item T-subst.
+\item[Tr-subst]
 Suppose $\sisubst{r}{S_1}{S_2} : \strin{s}$ for some $x$. 
@@
 and $\tsubst{S_1}{S_2} \err$ by the propagation rules corresponding to T-T-Subst.
-\end{itemize}
+
+\item[Tr-Coerce-Ok] TODO-nrf.
+
+\item[Tr-Coerce-NotOk] TODO-nrf.
+
+\end{description}
 \end{proof}
 
-\subsection{Type Safety Proof}
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+% END SECTION TWO -- THIS IS THE CONFLICT LINE
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 
-\subsection{New Theorems}
-
-\begin{thm}[Progress for Internal Language]
-\end{thm}
-
-\begin{thm}[Preservation for Internal Language]
-\end{thm}
-
-\begin{thm}[Representational Consistency]
-If $\cdot \vdash e : \tau \sreduces i$ then $\cdot \vdash i : \sigma$ and $\sigma \equiv \trep(\tau)$.
-\end{thm}
-
-\begin{thm}[Main Soundness Result]
-  If $\cdot \vdash e : \strin{r} \sreduces i$ then either $i \treduces \err$ or
-  if $i \sreduces v$ for $v$??? then $v \in r$.
-\end{thm}
-
-%\begin{thm}[Preservation]
-%  Let $T$ be a type in $\lcs$ and $h(T)=\sigma$ the corresponding type in the internal language.
-%  For all terms $e$:
-%  \begin{itemize}
-%    \item If $e:T$ and $e:T \treduces i$ then $i : \sigma$ such that $h(T) = \sigma$.
-%    \item If $e:T$ and $e:T \reduces e'$ then $e':T$.
-%  \end{itemize}
-%\end{thm}
-%\begin{proof}
-%The proof is a straightforward induction on the derivation of the combined evaluation relation.
-%\begin{itemize}[label=$ $,itemsep=1ex]
-%  \item \textbf{E-Ival, E-Ifilterval}. Both cases hold since the terms at hand are not $\lcs$ terms.
-%  \item \textbf{E-Stringval, E-strval}. Both cases hold since no reduction is possible. 
-%  \item \textbf{E-String}. By the definition of typing for internal terms, $\tstrf{e} : istr$. It suffices to show that $h(\str) = istr$, which follows from the definition of $h$.
-%  \item \textbf{E-String\_in}.
-By the definition of typing for internal terms, $\tstrf{e} : istr$.
-It suffices to show that $h(\strin[r]) = istr$, which follows from the definition of $h$ for arbitrary $r$.
-%  \item \textbf{E-concatval}.
-By the definition of typing for internal terms, $\tstrf{e_1} + \tstrf{e_2} : istr$.
-So it suffices to show that $h(\strin[r_1+r_2]) = istr$, which follows from the definition of $h$ for arbitrary $r_1$, $r_2$.
-%  \item \textbf{E-concatR, E-concatL}.
-Consider E-concatR.
-By induction, $e_1' : \strin{l_1}$.
-By inversion of T-Concat at the premise, $e_2 : \strin{r_2}$.
-Therefore, $e_1 + e_2 : \strin[r_1 + r_2]$.
-The left rule is symmetric.
-%  \item \textbf{E-Filterval}. We have that $\filter{\strin[r']}{e}:\strin[r\backslash r' + \emptyset]$. By inversion of T-Filter, $e:\strin[r]$.
-%    By T-Equiv-string\_in (which is bidirectional), $e \in \lang{r}$.
-%    By the Coercion Theorem, ${\tt rl\_filter}(r,e) \in \lang{r\backslash r' + \emptyset}$.
-%    By T-Equiv-string\_in,  $e \in \lang{r}$ and ${\tt rl\_filter}(r,e) \in \lang{r\backslash r' + \emptyset}$ implies ${\tt rl\_filter}(r,e):\strin[r\backslash r' + \emptyset]$.
-%  \item \textbf{E-Filter}. By inversion, $e : \strin[r']$ $\reduces e'$ so by the induction $e':\strin[r']$.
-%    Therefore, \\${\tt filter}(\strin[r], e'):$ $\strin[r \backslash r' + \emptyset]$ by T-Filter.
-%  \item \textbf{E-Convertval}. It suffices to show that $h(\strin[r]) = istr$, which is true by definition.
-%  \item \textbf{E-Convert}. By inversion and induction, $e' : \strin[r]$. We know that $[\strin[r']](e) : \strin[r']$, so by inversion of T-Convert $\lang{r'} \subseteq \lang{r}$.
-%    It follows that $[\strin[r']](e') : \strin[r']$.
-%  \item \textbf{E-DConvert}. By inversion, $e : \strin[r'] \reduces e'$. By the induction hypothesis, $e' : \strin[r']$; therefore, by T-Dconvert, $\dconvert[\strin[r]](e):\strin[r]$.
-%  \item \textbf{E-DConvertval}.
-By the definition of typing for internal terms, $\ifilter(r,\tstrf{e}) : istr$.
-It suffices to show that $h(\strin[r_1+r_2]) = istr$, which follows from the definition of $h$.
-%\end{itemize}
-%\end{proof}
-%
-%\begin{thm}[Progress]
-%  If $e:T$ then $e:T \reduces^* e' \treduces^* i$ where $i \ival$.
-%\end{thm}
-%\begin{proof}
-%By induction on the derivation of $e:T$.
-%For \textbf{T-Equiv-Include}, note that $e \in \lang{r}$, $e = \strf{s}$ for some $s$; therefore, $e \val$ by E-Strinval. 
-%We have that $e \val$ and $e : \strin[r]$, so $e \treduces \tstrf{e}$ by E-string\_in.
-%The remaining cases follow by induction on the hypotheses and application of corresponding evaluation rules.
-%\end{proof}
-%
-\section{Implementation Inside a Python}
-%
-%TODO-nrf rewrite
-%The $\lcs$ language is implemented as a type system extension in Ace; this extension
-%is illustrated in the examples at the beginning of this paper.
-%
-%Computing a regular expression representating the language
-%$R \backslash S$ is necessary in order to type-check expressions in which the filter function
-%occurs. This language is computed by translating $R$ and $S$ into finite automata,
-%complementing the final states of $S$, then constructing the cross-product of $R$
-%and $S'$. Type checking terminates only because this construction is decidable.
-%
-%In addition to this construction, other more typical operations -- such
-%as equality checks for regular expressions and regular expression matching -- are
-%also necessary. For these reasons, the Ace extension implementing $\lcs$ includes a library
-%implementing each of these constructions.
 \newcommand{\F}[1]{\textsf{#1}~}

### 2014-04-16 sanitation-plas14/sanitation-plas14.tex
More progess on Sec. 3

--- 2014-04-16 sanitation-plas14/sanitation-plas14.tex
@@
     \makeatletter
-
+%
 % \btIfInRange{number}{range list}{TRUE}{FALSE}
@@
 \newcommand{\Lagr}{\mathcal{L}}
+\newcommand{\strin}{{\tt string\_in}}
 \newcommand{\lang}[1]{\Lagr\{#1\}}
 \newcommand{\stru}[2]{ {\tt string\_union}(#1,#2)}
+
+\newcommand{\str}{ {\tt string} }
 
@@
 \newcommand{\reduces}{ \Rightarrow }
+\newcommand{\ireduces}{ \Downarrow }
+\newcommand{\sreduces}{ \leadsto }
 \newcommand{\gvd}{\Gamma \vdash }
@@
 
-\newcommand{\tstrf}[1]{`#1\textrm'} %??
+\newcommand{\val}{ \ {\tt val} }
+\newcommand{\ival}{ \ {\tt ival} }
+
+\newcommand{\istrf}[1]{`#1\textrm'} %??
 \newcommand{\strf}[1]{``#1"}
 
-%%
-%% Source and Target language definitions.
-%%
-% Source language terms.
-\newcommand{\sisubst}[3]{{\tt subst}[#1](#2,#3)}
-\newcommand{\coerce}[2]{ {\tt coerce}[#1](#2)}
-\newcommand{\sistr}[1]{{\tt si\_str}[#1]}
-\newcommand{\strin}[1]{\sistr{#1}}
-\newcommand{\siconcat}[2]{{\tt siconcat}(#1,#2)}
-
-% Source language types.
-\newcommand{\stringin}[1]{{\tt string\_in}[#1]}
-
-% target language terms.
-\newcommand{\tsubst}[3]{{\tt tsubst}(#1,#2,#3)}
-\newcommand{\tcheck}[2]{{\tt tcheck}(#1, #2)}
-\renewcommand{\tstr}[1]{{{\tt tstr}[#1]}}
-\newcommand{\tconcat}[2]{{\tt tconcat}(#1,#2)}
-
-% Target language types
-\newcommand{\str}{{\tt string}}
-\newcommand{\regext}[1]{ {\tt rx}[#1] }
-\newcommand{\rx}[1]{ {\tt rx}[#1] }
-
-% Meta-theoretic functions
-\newcommand{\dosubst}[3]{{\tt dosubst}(#1,#2,#3)}
-\newcommand{\lsubst}[3]{{\tt lsubst}(#1,#2,#3)}
-
-% Judgements
-\newcommand{\err}{{\tt err}}
-\newcommand{\trden}[1]{\llbracket #1 \rrbracket} % = Translation Denotation.
-
-% Relations
-\newcommand{\treduces}{ \Downarrow }
-\newcommand{\sreduces}{ \Downarrow }
-
-%%
-%% Constrain the size of full-page diagrams and rule lists
-%%
-%%\newcommand{\pagewidth}{5in}
-%%\newcommand{\rulelistwidth}{3in}
-
+%
+% Constrain the size of full-page diagrams and rule lists
+%
+%\newcommand{\pagewidth}{5in}
+%\newcommand{\rulelistwidth}{3in}
+
+%
 % Names of type systems presented in paper
+%
 \newcommand{\lcs}{\lambda_{CS}}
@@
 \setlength{\grammarindent}{3em}
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 
@@
 The semantics are defined in terms of an internal language with at least strings and a regex filter function.
-These constraints are captured by the internal term valuations.
+These constraints are captured by the internal term valuations ($\ival$).
 The internal language does not necessarily need a regex filter function because
@@
 
-% Whatever just dropping this in somewhere random.
-\begin{thm}
-If $S : \strin{r}$ then there exists a T such that $\trden{s} = T$ and either: 
-\begin{itemize}
-\item (a) $T \treduces \tstr{s}$ and $S \sreduces \strin{s}$, and $s \in lang{r}$.
-\item (b) $T \err$ and $S \err$.
-\end{itemize}
-\end{thm}
-\begin{proof}
-By induction on the derivation of $\trden{S} = T$. TODO-nrf add case magic.
-\begin{itemize}
-\item Tr-string. By assumption $\trden{S} = \tstr{s}$. The (a) property follows.
-By S-E-Str, $S \sreduces \str{s}$ and by T-E-Str, $\tstr{s} \treduces \tstr{s}$.
-By source language type safety, it follows that $s \in \lang{r}$.
-\item Tr-Concat.
-By inversion of the typing relaiton for the source lnaguage, $S_1 : \strin{r_1}$ and $S_2 : \stringin{S_2}$.
-By induction on the left premise, $T_1 \treduces \tstr{s_1}$, $S_1 \sreduces \str{s_1}$ and $s_1 \in \lang{r_1}$
-or else $T_1 err$. 
-Similarly for $T_2$ and $S_2$.
-In the error cases, the propagations rules corresponding to S-E-Concat and T-E-Concat imply that $\tconcat{T_1}{T_2} \err$ and $\siconcat{S_1}{S_2} \err$.
-In the non-error case, $\siconcat{S_1}{S_2} \sreduces \strin{s_1 \cdot s_2}$ by S-E-Concat and $\tconcat{T_1}{T_2} \treduces \tstr{s_1 \cdot s_2}$ by T-E-Concat.
-Finally, $s_1 \in \lang{r_1}$ and $s_2 \in \lang{r_2}$ implies that $s_1 \cdot s_2 \in \lang{r}$ by \ref{lemmaregex}.
-\item T-subst.
-Suppose $\sisubst{r}{S_1}{S_2} : \strin{s}$ for some $x$. 
-By inverison, $S_1 : \strin{r_1}$ and $S_2 : \strin{r_2}$.
-Note that $s : \subst{r}{S_1}{S_2} : \stringin{\tsubst{r}{r_1}{r_2}}$ by S-T-Subst.
-By induction, $\trden{S_1} = T_1$ and $\trden{S_2} = T_2$.
-Either (a) or (b) holds for each.
-
-If (a) holds for both, the $S_1 \sreduces \strin{S_1}$ and $S_2 \sreduces \strin{S_2}$ for $s_1 \in \lang{r_1}$ and $s_2 \in \lang{r_2}$.
-Therefore, $\sisubst{r}{S_1}{S_2} \sreduces \strin{\dosubst{r,s_1,s_2}}$.
-The same fact can be shown in an analogous manner for $T_1$ and $T_2$.
-Finally, $s_1 \in \lang{r_1}$ and $s_2 \in \lang{r_2}$, which implies that $\dosubst{r}{s_1}{s_2} \in \lang{\sisubst{r}{s_1}{s_2}}$ by \ref{substhm}.
-
-If (b) holds for either the left or right premise, then $\sisubst{r}{S_1}{S_2} \err$ by the propagation rules corresponding to S-T-subst, 
-and $\tsubst{S_1}{S_2} \err$ by the propagation rules corresponding to T-T-Subst.
-\end{itemize}
-\end{proof}
-
 %
@@
 %
+\newcommand{\ereplace}[3]{{\tt replace}[#1](#2,#3)}
+\newcommand{\ecoerce}[2]{{\tt coerce}[#1](#2)}
+\newcommand{\estr}[1]{{\tt str\_in[#1]}}
+\newcommand{\estrt}[1]{{\tt string\_in[#1]}}
+\newcommand{\econcat}[2]{{\tt concat}(#1,#2)}
+
+\newcommand{\ireplace}[3]{{\tt ireplace}(#1,#2,#3)}
+\newcommand{\icheck}[2]{{\tt icheck}(#1, #2)}
+\newcommand{\istr}[1]{{{\tt istr}[#1]}}
+\newcommand{\istrt}{{\tt istrt}}
+\newcommand{\iconcat}{\hat{+}}
+\newcommand{\regext}{ {\tt regex} }
+
+% Special stuff for the semantics.
+\newcommand{\replacefn}[3]{\tt replacement(#1,#2,#3)}
+\newcommand{\crl}[3]{{\tt compute\_replacement\_lanugage}(#1,#2,#3)}
 
@@
 
-<$\psi$> ::=				\hfill	Source language types					\\
-$\stringin{r}$				\hfill Regular expression types \alt
-
-<S> ::= \hfill Source language expressions \\
-      $\strin{S}$ \alt
-      $\siconcat{S}{S}$ \hfill String concatenation \alt
-      $\sisubst{r}{S}{S}$ \alt
-      $\coerce{r}{S}$
-
-<$\theta$> ::= \hfill Target language types \\
-$\str$ \alt $\regext$
-
-<P> ::= \hfill Internal expressions \\
-  $\tstr{P}$ \alt
-  $\siconcat{P}{P}$ \alt
-  $\tsubst{r}{P}{P}$ \alt
-  $\tcheck{r}{P}$
+<$\psi$> ::=				\hfill	External types					\\
+$\estrt{r}$				\hfill Regular expression types \alt
+
+<T> ::= \hfill External expressions \\
+      $\estr{s}$ \alt
+      $\econcat{e}{e}$ \hfill String concatenation \alt
+      $\ereplace{r}{e}{e}$ \alt
+      $\ecoerce{r}{e}$
+
+<$\theta$> ::= \hfill Internal types \\
+$\istrt$ \alt $\regext$
+
+<S> ::= \hfill Internal expressions \\
+  $\istr{s}$ \alt
+  $\econcat{i}{i}$ \alt
+  $\ireplace{r}{i}{i}$ \alt
+  $\icheck{r}{i}$
 \end{grammar}
@@
 %
-%There are two evaluation judgements: $e:T \reduces e'$ and $e:T \treduces i$.
-%The $\reduces$ relation is between $\lcs$ expressions, while the $\treduces$
+%There are two evaluation judgements: $e:T \reduces e'$ and $e:T \ireduces i$.
+%The $\reduces$ relation is between $\lcs$ expressions, while the $\ireduces$
 %relation is a mapping from $\lcs$ expressions into internal language expressions
@@
 %\begin{defn}[Types of internal values]
-%  Let $\tstrf{s}$ range over string literals and $r$ over regular expressions.
+%  Let $\istrf{s}$ range over string literals and $r$ over regular expressions.
 %  Internal values are typed as follows:
 %\begin{itemize}
-%  \item If $e=\tstrf{s}$ then $e:istr$.
-%  \item If $e=\ifilter{r}{\tstrf{s}}$ then $e:istr$.
-%  \item If $e=\tstrf{s_1}+\tstrf{s_2}$ then $e:istr$.
+%  \item If $e=\istrf{s}$ then $e:istr$.
+%  \item If $e=\ifilter{r}{\istrf{s}}$ then $e:istr$.
+%  \item If $e=\istrf{s_1}+\istrf{s_2}$ then $e:istr$.
 %\end{itemize}
@@
 \onecolumn
-\newcommand{\sctx}{\Gamma} % Context for external typing
-\newcommand{\tctx}{\Omega} % Context for internal typing
+\newcommand{\ectx}{?} % Context for external typing
+\newcommand{\ictx}{?} % Context for internal typing
 
@@
 
-\inferline{S-T-include}
-{\sctx \vdash \strin{s} : \stringin{r}}
-{s \in \lang{r}}
-
-\inferline{S-T-Concat}
-{\sctx \vdash \siconcat{S_1}{S_2} : \strin{r_1 \cdot r_2}}
-{\sctx \vdash S_1 : \strin{r_1} \\ \sctx \vdash S_2 : \strin{r_2}}
-
-\inferline{S-T-subst}
-{\sctx \vdash \sisubst{r}{S_1}{S_2} : r'}
-{\sctx \vdash S_1 : \strin{r_1} \\ \sctx \vdash S_2 : \strin{r_2} \\ {\tt lsubst}(r,r_1,r_2) = r'}
-
-\inferline{S-T-coerce}
-{\sctx \vdash \coerce{r'}{s} : \stringin{r'}}
-{\sctx \vdash S : \stringin{r}}
+\inferline{Te-include}
+{\ectx \vdash \estr{s} : \estrt{r}}
+{s \in \lang{r}???}
+
+\inferline{Te-Concat}
+{\ectx \vdash \econcat{S_1}{S_2} : \estrt{r_1 \cdot r_2}}
+{\ectx \vdash S_1 : \estrt{r_1} \\ \ectx \vdash S_2 : \estrt{r_2}}
+
+\inferline{Te-subst}
+{\ectx \vdash \ereplace{r}{S_1}{S_2} : r}
+{\ectx \vdash S_1 : \estrt{r_1} \\ \ectx \vdash S_2 : \estrt{r_2} \\ r = {\tt lsubst}(r,r_1,r_2)}
+
+\inferline{Te-coerce}
+{\ectx \vdash \ecoerce{r'}{s} : \estrt{r'}}
+{\ectx \vdash S : \estrt{r}}
 
 \end{tabular}
-\caption{Typing rules for source langauge terms (S). The metavariable s ranges over
-string literals.}
+\caption{Typing rules for $\lcs$. Err cases?}
 \label{fig:etyping}
@@
 \begin{tabular}{c r}
-\inferline{S-E-Include}
-{\strin{s} \sreduces \strin{s}}
+\inferline{Te-Include}
+{\estr{s} \val}
 { \cdot }
 
-\inferline{S-E-Concat}
-{\siconcat{S_1}{S_2} \sreduces \strin{s_1 \cdot s_2}} % ???
-{S_1 \sreduces \strin{s_1} \\ S_2 \sreduces \strin{s_2}}
-
-\inferline{S-E-Subst}
-{\sisubst{r}{S_1}{S_2} \sreduces s}
-{S_1 \sreduces \strin{s_1} \\ S_2 \sreduces \strin{s_2} \\ \dosubst{r}{s_1}{s_2} = s}
-
-\inferline{S-E-Coerce-OK}
-{\coerce{r}{S} \sreduces \strin{s}}
-{S \sreduces \strin{s} \\ x \in \lang{r}}
-
-\inferline{S-E-Coerce-NotOk}
-{S \err}
-{\coerce{r}{S} \sreduces \strin{s} \\ s \not \in \lang{r}}
-
+\inferline{Te-Concat}
+{\econcat{S_1}{S_2} \ereduces \estr{s_1 \cdot s_2}} % ???
+{S_1 \Downarrow \estr{s_1} \\ S_2 \Downarrow \estr{s_2}}
+
+\inferline{Te-Subst}
+{\ereplace{r}{S_1}{S_2} \ereduces s}
+{S_1 \Downarrow \estr{s_1} \\ S_2 \Downarrow \estr{s_2} \\ s = {\tt substitution}(r,s_1,s_2)}
+
+\inferline{Te-Coerce-OK}
+{\ecoerce{r}{s} \ereduces \estr{x}}
+{s \ereduces \estr{x} \\ x \in \lang{r}}
+
+\inferline{Te-Coerce-NotOk}
+{\ecoerce{r}{s} \ereduces \estr{x} \\ x \not \in \lang{r}}
+{s \ereduces \estr{x} \\ x \not \in \lang{r}}
 \end{tabular}
-\caption{Big step semantics for the string fragment of the source language.
-There are also unmentioned error propagation rules for for S-E-Concat and S-E-Subst.}
+\caption{Operational semantics for the external language $\lcs$.
+There are also unmentioned error propagation rules for for Te-Concat and Te-Subst.}
 \label{fig:eeval}
@@
 \begin{tabular}{c r}
-
-\inferline{P-T-String}
-{\tctx \vdash \tstr{s} : \str}
-{\cdot}
-
-\inferline{P-T-Concat}
-{\tctx \vdash \tconcat{P_1}{P_2} : \str}
-{\tctx \vdash P_1 : \str \\ \tctx \vdash P_2 : \str}
-
-\inferline{P-T-Subst}
-{\tctx \vdash \tsubst{\rx{r}}{P_1}{P_2} : \str}
-{\tctx \vdash P_1 : \str \\ \tctx \vdash P_2 : \str}
-
-\inferline{P-T-Check}
-{\tctx \vdash \tcheck{\rx{r}}{s} : \str}
-{\tctx \vdash P : \str}
-
+\inferline{Ti-Include}
+{\istr{s} \val}
+{ \cdot }
+
+\inferline{Ti-Include}
+{{\tt regex}[r] \val}
+{ \cdot }
+
+\inferline{Ti-Concat}
+{\iconcat{S_1}{S_2} \ireduces \istr{s_1 \cdot s_2}} % ???
+{S_1 \ireduces \istr{s_1} \\ S_2 \ireduces \istr{s_2}}
+
+\inferline{Te-Subst}
+{\ireplace{r}{S_2}{S_3} \ireduces {\tt substitution(r,s_2,s_3)}}
+{T_1 \ireduces {\tt regex}[r] \\ T_2 \ireduces \istr{s_2} \\ T_3 \ireduces \istr{s_3} \\ }
+
+\inferline{Te-Ccheck-OK}
+{\icheck{r}{s} \ireduces \istr{x}}
+{s \ireduces \estr{x} \\ x \in \lang{r}}
+
+\inferline{Te-Check-NotOk}
+{\icheck{r}{s} \ireduces \istr{x} \\ x \not \in \lang{r}}
+{s \ireduces \istr{x} \\ x \not \in \lang{r}}
 \end{tabular}
-\caption{Typing rules for target langauge terms (P). The metavariable s ranges over
-string literals.}
-\label{fig:etyping}
+\caption{Operational semantics for the internal language of $\lcs$.
+There are also unmentioned error propagation rules for for Te-Concat and Te-Subst.}
+\label{fig:ieval}
 \end{center}
@@
 
+%
+%\begin{figure}
+%\begin{center}
+%\begin{tabular}{c r}
+%
+%\inferline{T-Abs}
+%{\gvd \lambda x : \tau .e : \tau \rightarrow \tau' \sreduces \lambda x : \sigma.i}
+%{\Gamma, x : \tau \vdash e : \tau' \sreduces i \\ \sigma \equiv \trep(\tau)}
+%
+%\inferline{T-App}
+%{\Gamma \vdash e_1 e_2 \sreduces i_1 i_2}
+%{\Gamma \vdash e_1 : \tau \rightarrow \tau' \sreduces i_1 \\ \Gamma \vdash e_2 : \tau \sreduces i_2}
+%
+%\inferline{T-Replace}
+%{\gvd \ereplace{r}{e_1}{e_2} : \estrt{r'} \sreduces \ireplace{r}{i_1}{i_2}}
+%{\gvd e_1 : \estrt{r_1} \ireduces i_1 \\ \gvd e_2 : \estrt{r_2} \ireduces i_2 \\ r' \equiv \crl{r}{r_2}{r_1}}
+%
+%\inferline{T-Check-Ok}
+%{\gvd \echeck{r'}{e} : \estrt{r'} \sreduces i}
+%{\gvd e : \estrt{r} \sreduces{i} \\ \lang{r'} \subseteq \lang{r}}
+%
+%\inferline{T-Check-NotOk}
+%{\gvd \echeck{r'}{e} : \estrt{r'} \sreduces \icheck{r}{i}}
+%{\gvd e : \estrt{r} \sreduces i \\ \lang{r'} \not \subseteq \lang{r}}
+%  
+%%\inferline{E-dconvert (optional) }
+%%{\dconvert{\strin[r']}{e} : \strin[r'] \ireduces \ifilter{r'}{\istrf{e}}}
+%%{e \val }
+%
+%\end{tabular}
+%\caption{Type-directed translation to internal terms for $\lcs$}
+%\label{fig:eeval}
+%\end{center}
+%\end{figure}
+
+%
+% Begin typing relation for internal terms.
+%
 \begin{figure}
@@
 \begin{tabular}{c r}
-\inferline{P-E-Str}
-{\tstr{s} \treduces \str{s}}
-{ \cdot }
-
-\inferline{P-E-Include}
-{\rx{r} \treduces \rx{r}}
-{ \cdot }
-
-\inferline{P-E-Concat}
-{\tconcat{P_1}{P_2} \treduces \tstr{s_1 \cdot s_2}} % ???
-{P_1 \treduces \tstr{s_1} \\ P_2 \treduces \tstr{s_2}}
-
-\inferline{P-E-Subst}
-{\tsubst{r}{P_2}{P_3} \treduces \tstr{s}}
-{P_1 \treduces \rx{r} \\ P_2 \treduces \tstr{s_2} \\ P_3 \treduces \tstr{s_3} \\ \dosubst{r}{s_2}{s_3} = s}
-
-\inferline{P-E-Check-OK}
-{\tcheck{r}{P} \treduces \tstr{s}}
-{P \treduces \strin{s} \\ s \in \lang{r}}
-
-\inferline{P-E-Check-NotOk}
-{P \err}
-{\tcheck{r}{T} \treduces \tstr{s} \\ s \not \in \lang{r}}
+\inferline{Ti-Var}
+{\ovd x : \sigma}
+{x : \sigma \in \Omega}
+
+% Add in pedantic rules as well? If there's space.
+\inferline{Ti-Abs}
+{\ovd \lambda x : \sigma . i : \sigma \rightarrow \sigma'}
+{\Omega, x : \sigma \vdash i : \sigma'}
+
+\inferline{Ti-App}
+{i_1 i_2 : \sigma'}
+{\ovd i_1 : \sigma \rightarrow \sigma' \\ \ovd i_2 : \sigma}
+
+\inferline{Ti-Str}
+{\ovd \istr{s} : \istrt }
+{\cdot}
+
+\inferline{Ti-Replace}
+{\ovd \ireplace{r}{i_1}{i_2} : \istrt}
+{\ovd i_1 : \istrt \\ \ovd i_2 : \istrt}
+
+\inferline{Ti-Check}
+{\ovd \icheck{r}{i_1}{i_2} : \istrt}
+{\ovd i_1 : \istrt \\ \ovd i_2 : \istrt}
+
+\inferline{Ti-Concat}
+{\ovd i_1 \iconcat i_2 : \istrt}
+{\ovd i_1 : \istrt \\ \ovd i_2 : \istrt}
 
 \end{tabular}
-\caption{Big step semantics for the target lanugage.
-There are also unimplemented error propagation rules for for Te-Concat and Te-Subst.}
-\label{fig:ieval}
+\caption{Typing relation on internal terms.}\label{fig:itypes}
 \end{center}
@@
 %
-
+\newcommand{\err}{{\tt err}}
 \begin{figure}
@@
 
-\inferline{Tr-string}
-{\trden{\strin{s}} = \tstr{s}}
-{ \ }
-
-\inferline{Tr-Concat}
-{\trden{\siconcat{S_1}{S_2}} = \tconcat{P_1}{P_2}}
-{ \trden{S_1} = P_1 \\ \trden{S_2} = P_2}
-
-\inferline{Tr-Subst}
-{\trden{ \sisubst{r}{S_1}{S_2} } = \tsubst{\rx{r}}{P_1}{P_2} }
-{ \trden{S_1} = P_1 \\ \trden{S_2} = P_2 }
-
-\inferline{Tr-Subst}
-{\trden{ \coerce{r'}{S} } = P}
-{ \trden{s} = P \\ S : \strin{r} \\ \lang{r'} \subseteq \lang{r}}
-
-\inferline{Tr-Coerce-NotOk}
-{\trden{\coerce{r'}{S}} = \tcheck{\rx{r'}}{S}}
-{\trden{S} = P \\ s : \strin{r} \\ \lang{r'} \not \subseteq \lang{r}}
+  \inferline{Ei-strval}
+  {\istr{s} \ival}
+  { \cdot }
+
+  \inferline{Ei-Replace}
+  {\ireplace{r}{i_1}{i_2} \ireduces i'}
+  { i \ireduces \replacefn{r}{i_1}{i_2} }
+
+  \inferline{Ei-CheckOk}
+  {\icheck{r}{i} \ireduces i}
+  {i \in \lang{r}}
+
+  \inferline{Ei-Check-NotOk}
+  {\icheck{r}{i} \ireduces \err}
+  {i \not \in \lang{r}}
+
+  \inferline{Ei-Concat}
+  { i_1 \iconcat i_2 \ireduces \istr{i_1 \iconcat i_2} }
+  { \cdot }
 
 \end{tabular}
-\caption{Translation from source terms (S) to target terms (P).
-The translation is type-directed in the Tr-Coerce cases.}
+\caption{Big-step semantics for internal language; standard rules for the STL are omitted for brevity. The meta-variable $s$ ranges over string literals.}\label{fig:ieval}
 \end{center}
@@
 that a check must succeed. Furthermore, expensive calculations (such as language
-inclusion or ${\tt dsubst}$, all occur at compile time.
+inclusion or ${\tt computer\_replacement\_language}$, all occur at compile time.
 Since sanitation problems are generally expressible with smalle, simple expressions,
-we believe that the compile-time overhead is not significant enough to prohibiour target usecase. Informal experimentation with our implementation seems to
+we believe that the compile-time overhead is not significant enough to prohibit
+our target usecase. Informal experimentation with our implementation seems to
 support this assertion.
@@
 \begin{thm}[Main Soundness Result]
-  If $\cdot \vdash e : \strin{r} \sreduces i$ then either $i \treduces \err$ or
-  if $i \sreduces v$ for $v$??? then $v \in r$.
+  If $\cdot \vdash e : \estrt{r} \sreduces i$ then either $i \ireduces \err$ or
+  if $i \sreduces v$ for $v \ival$ then $v \in r$.
 \end{thm}
@@
 %  \begin{itemize}
-%    \item If $e:T$ and $e:T \treduces i$ then $i : \sigma$ such that $h(T) = \sigma$.
+%    \item If $e:T$ and $e:T \ireduces i$ then $i : \sigma$ such that $h(T) = \sigma$.
 %    \item If $e:T$ and $e:T \reduces e'$ then $e':T$.
@@
 %  \item \textbf{E-Stringval, E-strval}. Both cases hold since no reduction is possible. 
-%  \item \textbf{E-String}. By the definition of typing for internal terms, $\tstrf{e} : istr$. It suffices to show that $h(\str) = istr$, which follows from the definition of $h$.
+%  \item \textbf{E-String}. By the definition of typing for internal terms, $\istrf{e} : istr$. It suffices to show that $h(\str) = istr$, which follows from the definition of $h$.
 %  \item \textbf{E-String\_in}.
-By the definition of typing for internal terms, $\tstrf{e} : istr$.
+By the definition of typing for internal terms, $\istrf{e} : istr$.
 It suffices to show that $h(\strin[r]) = istr$, which follows from the definition of $h$ for arbitrary $r$.
 %  \item \textbf{E-concatval}.
-By the definition of typing for internal terms, $\tstrf{e_1} + \tstrf{e_2} : istr$.
+By the definition of typing for internal terms, $\istrf{e_1} + \istrf{e_2} : istr$.
 So it suffices to show that $h(\strin[r_1+r_2]) = istr$, which follows from the definition of $h$ for arbitrary $r_1$, $r_2$.
@@
 %  \item \textbf{E-DConvertval}.
-By the definition of typing for internal terms, $\ifilter(r,\tstrf{e}) : istr$.
+By the definition of typing for internal terms, $\ifilter(r,\istrf{e}) : istr$.
 It suffices to show that $h(\strin[r_1+r_2]) = istr$, which follows from the definition of $h$.
@@
 %\begin{thm}[Progress]
-%  If $e:T$ then $e:T \reduces^* e' \treduces^* i$ where $i \ival$.
+%  If $e:T$ then $e:T \reduces^* e' \ireduces^* i$ where $i \ival$.
 %\end{thm}
@@
 %For \textbf{T-Equiv-Include}, note that $e \in \lang{r}$, $e = \strf{s}$ for some $s$; therefore, $e \val$ by E-Strinval. 
-%We have that $e \val$ and $e : \strin[r]$, so $e \treduces \tstrf{e}$ by E-string\_in.
+%We have that $e \val$ and $e : \strin[r]$, so $e \ireduces \istrf{e}$ by E-string\_in.
 %The remaining cases follow by induction on the hypotheses and application of corresponding evaluation rules.
@@
 \newcommand{\Q}{\FF{Arg}}
-\newcommand{\xlA}[1]{\lfloor #1 \rfloor_{\lamAce}}
-\newcommand{\xA}[1]{$\lfloor #1 \rfloor_{\texttt{ace}}$}
-\begin{figure*}
-\begin{tabular}{ l r l l }
-$\tau$ & & $\xlA{\tau}$ & \xA{\tau}\\
+\newcommand{\xlA}[1]{\lfloor #1 \rfloor_{\lambda}}
+\newcommand{\xA}[1]{$\lfloor #1 \rfloor_{\text{Ace}}$}
+\newcommand{\rtotau}[1]{\lfloor #1 \rfloor}
+
+\newcommand{\tremp}{{\tt tremp}}
+\newcommand{\trdot}{{\tt trdot}}
+\newcommand{\trchar}[1]{{\tt trchar}[#1]}
+\newcommand{\trseq}[2]{{\tt trseq}(#1; #2)}
+\newcommand{\tror}[2]{{\tt tror}(#1; #2)}
+\newcommand{\tlstr}[1]{{\tt tstr}[#1]}
+\begin{table*}[t]
+\centering
+\begin{tabular}{ l l l }
+$r$ & $\xlA{r}$ & \xA{r}\\
 \hline
-$\strin{r}$ & & $\fvar{stringin}[r]$ & \verb|string_in["|$r$\verb|"]|\\
+$\epsilon$ & $\tremp$ & \verb|""|\\
+$.$ & $\trdot$ & \verb|"."|\\
+$a$ & $\trchar{a}$ & \verb|"|$a$\verb|"|\\
+$r_1 \cdot r_2$ & $\trseq{\xlA{r_1}}{\xlA{r_2}}$ & \xA{r_1}\verb| + |\xA{r_2}\\
+$r_1 + r_2$ & $\tror{\xlA{r_1}}{\xlA{r_2}}$ & \verb|"(" + |\xA{r_1}\verb! + "|" + !\xA{r_2}\verb| + ")"|\\
 \\
-  $e$ & & $\xlA{e}$ & \xA{e}\\
+$\psi$ & $\xlA{\psi}$ & \xA{\psi}\\
 \hline
-  ${\tt str}[r, s]$ & \text{synthetic position} & $\FF{intro}[\FF{str}[s]]()$ : \fvar{stringin}[\FF{rx}[$r$]] & \verb|"|$s$\verb|" (string_in["|$r$\verb|"])|\\
-  & \text{analytic position} & $\FF{intro}[\FF{str}[s]]()$ & \verb|"|$s$\verb|"|\\
-  ${\tt concat}(e_1; e_2)$ & & $\xlA{e_1}\cdot\FF{elim}[\tvar{concat}](\xlA{e_2})$ & \xA{e_1}\verb| + |\xA{e_2} \\
-  ${\tt replace}[r](e_1; e_2)$ & & $\xlA{e_1}\cdot\FF{elim}[\tvar{replace}~\FF{rx}[r]](\xlA{e_2})$ & \xA{e_1}\verb|.replace("|$r$\verb|", |\xA{e_2}\verb|)|\\
-  ${\tt coerce}[r](e)$ & & $\xlA{e}\cdot\FF{elim}[\tvar{coerce}~\FF{rx}[r]]()$ & \xA{e}\verb|.coerce("|$r$\verb|")|
+${\tt string\_in}[r]$ & $\fvar{stringin}[\xlA{r}]$ & \verb|string_in[|\xA{r}\verb|]|\\
+\\
+  $S$ & $\xlA{S}$ & \xA{S} \\
+  \hline
+  ${\tt str\_in}[s]$ & $\FF{intro}[\FF{str}[s]]()$ & \verb|"|$s$\verb|"|\\
+  ${\tt concat}(S_1; S_2)$  & $\xlA{S_1}\cdot\FF{elim}[\tvar{concat}](\xlA{S_2})$ & \xA{S_1}\verb| + |\xA{S_2} \\
+  ${\tt subst}[r](S_1; S_2)$  & $\xlA{S_1}\cdot\FF{elim}[\tvar{subst}~\xlA{r}](\xlA{S_2})$ & \xA{S_1}\verb|.subst(|\xA{r}\verb|, |\xA{S_2}\verb|)|\\
+  ${\tt coerce}[r](S)$  & $\xlA{S}\cdot\FF{elim}[\tvar{coerce}~\xlA{r}]()$ & \xA{S}\verb|.coerce(|\xA{r}\verb|)|
 \end{tabular}
-\caption{Translation of $\lcs$ to $\lamAce$ and Ace.}
+\caption{Embeddings of the ${\tt string\_in}$ fragment into $\lamAce$ and Ace.}
+\end{table*}
+%
+\newcommand{\atjsynX}[3]{\Gamma \vdash_\fvalCtx #1 \Rightarrow #2 \leadsto #3}
+\newcommand{\atjanaX}[3]{\Gamma \vdash_\fvalCtx #1 \Leftarrow #2 \leadsto #3}
+\newcommand{\atjerrX}[1]{\Gamma \vdash_\fvalCtx #1~ \mathtt{error}}
+\begin{figure*}[t]
+\small
+$\fbox{\inferrule{}{\atjsynX{e}{\tau}{i}}}$~~~~
+$\fbox{\inferrule{}{\atjanaX{e}{\tau}{i}}}$~~~~
+%$\fbox{\inferrule{}{\atjerrX{e}}}$
+\begin{mathpar}
+\inferrule[att-flip]{
+	\atjsynX{e}{\tau}{i}
+}{
+	\atjanaX{e}{\tau}{i}
+}
+
+%\inferrule[att-var]{
+%	x \Rightarrow \tau \in \Gamma
+%}{
+%	\atjsynX{x}{\tau}{x}
+%}
+%
+%\inferrule[att-asc]{
+%    \tau \Downarrow_\fvalCtx \tau'\\
+%	\atjanaX{e}{\tau'}{i}
+%}{
+%	\atjsynX{e : \tau}{\tau'}{i}
+%}
+%
+%\inferrule[att-let-syn]{
+%	\atjsynX{e_1}{\tau_1}{i_1}\\
+%	\Gamma, x \Rightarrow \tau_1 \vdash_\fvalCtx e_2 \Rightarrow \tau_2 \leadsto i_2\\
+%	\trepof{\tau_1} \Downarrow_\fvalCtx \titype{\sigma_1}
+%}{
+%	\atjsynX{\F{let}x = e_1~\F{in}e_2}{\tau_2}{(\ilam{x}{\sigma_1}{i_2})~i_1}
+%}
+%
+%\inferrule[att-lam-ana]{
+%	\Gamma, x \Rightarrow \tau_1 \vdash_\fvalCtx e \Leftarrow \tau_2 \leadsto i\\
+%	\trepof{\tau_1} \Downarrow_\fvalCtx \titype{\sigma_1}
+%}{
+%	\atjanaX{\lambda x.e}{\ttype{arrow}{(\tau_1, \tau_2)}}{\ilam{x}{\sigma_1}{i}}
+%}
+%
+%\inferrule[att-lam]{
+%	\tau_1 \Downarrow_\fvalCtx \tau_1'\\
+%		\trepof{\tau_1'} \Downarrow_\fvalCtx \titype{\sigma}\\\\
+%	\Gamma, x \Rightarrow \tau_1' \vdash_\fvalCtx e \Rightarrow \tau_2 \leadsto i
+%%	i \hookrightarrow_\fvalCtx i'\\
+%%		\sigma \hookrightarrow_\fvalCtx \sigma'
+%%	\ddbar{\fvar{Arrow}}{\fvalCtx}{\trepof{\tau_1'}}{\sbar_1}\\
+%	%\delfromtau{$\Xi_0$}{\fvalCtx}{\tau_1'}{\sabs}\\\\
+%}{
+%	\atjsynX{\elam{x}{\tau_1}{e}}{\ttype{arrow}{(\tau_1', \tau_2)}}{\ilam{x}{\sigma'}{i}}
+%}
+%
+\inferrule[att-intro-ana]{
+	\vdash_\fvalCtx \FF{iana}(\fvar{tycon})=\taudef\\
+	\taudef~\taut{opidx}~\taut{tyidx}~((\Gamma; e_1)? :: \ldots :: (\Gamma; e_n)? :: []) \Downarrow_\fvalCtx \titerm{i}\\\\
+%	\trepof{\ttype{tycon}{\tauidx'}} \Downarrow_\fvalCtx \titype{\sigma}\\
+		\trepof{\ttype{tycon}{\taut{tyidx}}} \Downarrow_\fvalCtx \titype{\sigma}\\
+		\Gamma \vdash_\fvalCtx i : \sigma
+}{
+	\atjanaX{\FF{intro}[\taut{opidx}](e_1; \ldots; e_n)}{\ttype{tycon}{\taut{tyidx}}}{i}
+}
+
+%\inferrule[att-i-asc-ty]{
+%	\atjanaX{I}{\tau}{i}
+%}{
+%	\atjsynX{I : \tau}{\tau}{i}
+%}
+%
+%\inferrule[att-i-asc-tycon]{
+%	\vdash_\fvalCtx \FF{isyn}(\fvar{tycon})=\taudef\\\\
+%	\taudef~\tauidx~((\Gamma; e_1)? {::}{\ldots}{::}(\Gamma; e_n)? {::} []) \Downarrow_\fvalCtx \tden{\titerm{i}}{\ttype{tycon}{\tauidx'}}\\
+%			\trepof{\tau_1'} \Downarrow_\fvalCtx \titype{\sigma}\\
+%	\Gamma \vdash_\fvalCtx i : \sigma
+%}{
+%	\atjsynX{\FF{intro}[\tauidx](e_1; \ldots; e_n)] :: \fvar{tycon}}{\ttype{tycon}{\tauidx'}}{i'}
+%}
+%
+\inferrule[att-elim-syn]{
+	\atjsynX{e}{\ttype{tycon}{\taut{tyidx}}}{i}\\
+	\vdash_\fvalCtx \FF{esyn}(\fvar{tycon})=\taudef\\\\
+	\taudef~\taut{opidx}~\taut{tyidx}~\titerm{i}~((\Gamma; e)? :: (\Gamma; e_1)? :: \ldots :: (\Gamma; e_n)? :: []) \Downarrow_\fvalCtx (\tau, \titerm{i'})\\\\
+			\trepof{\tau} \Downarrow_\fvalCtx \titype{\sigma}\\
+	\Gamma \vdash_\fvalCtx i' : \sigma
+}{
+	\atjsynX{e\cdot\FF{elim}[\taut{opidx}](e_1; \ldots; e_n)}{\tau}{i'}
+}
+\end{mathpar}
+%\vspace{-10px}
+\caption{The bidirectional active typechecking and translation judgements.}
+\label{atj}
 \end{figure*}
-\begin{figure}
+\newcommand{\tlevalX}[2]{#1 \Downarrow_\fvalCtx #2}
+\begin{figure*}[t]
+\small
+$\fbox{\inferrule{}{\tau \Downarrow_\fvalCtx \tau'}}$~~~~
+\begin{mathpar}
+\inferrule[trstr-eval]{ }{\tlevalX{\tlstr{s}}{\tlstr{s}}}
+
+\inferrule[tremp-eval]{ }{\tremp \Downarrow_\fvalCtx \tremp}
+
+\cdots
+
+\inferrule[tror-eval]{
+	\tlevalX{\tau_1}{\tau_1'}\\
+	\tlevalX{\tau_2}{\tau_2'}
+}{
+	\tlevalX{\tror{\tau_1}{\tau_2}}{\tror{\tau_1'}{\tau_2'}}
+}
+
+trmatch
+
+trlsubst
+
+trsublang
+
+%\inferrule[repof]{
+%	\tau \Downarrow_\fvalCtx \ttype{tycon}{\tauidx}\\
+%	\vdash_\fvalCtx \FF{rep}(\fvar{tycon}) = \taurep\\
+%	\taurep~\tauidx \Downarrow_\fvalCtx \titype{\sigma}
+%}{
+%	\FF{repof}(\tau) \Downarrow_\fvalCtx \titype{\sigma}
+%}
+%
+\inferrule[syn]{
+    \tau \Downarrow_\fvalCtx (\Gamma; e)?\\
+	\atjsynX{e}{\tau}{\iota}\\\\
+    [\tau/\tvar{t}_{ty}, \titerm{\iota}/\tvar{t}_{trans}]\tau_2 \Downarrow_\fvalCtx \tau_2'
+}{
+	\FF{syn}(\tau_1; \tvar{t}_{ty}, \tvar{t}_{trans}.\tau_2) \Downarrow_\fvalCtx \tau_2'
+}
+
+\inferrule[ana]{
+	\tau_1 \Downarrow_\fvalCtx (\Gamma; e)?\\
+	\tau_2 \Downarrow_\fvalCtx \tau_2'\\
+	\atjanaX{e}{\tau_2'}{\iota}\\\\
+	[\titerm{\iota}/\tvar{t}_{trans}]\tau_3 \Downarrow_\fvalCtx \tau_3'
+}{
+	\FF{ana}(\tau_1; \tau_2; \tvar{t}_{trans}.\tau_3) \Downarrow_\fvalCtx \tau_3'
+}
+
+TODO: add rx, str stuff
+\end{mathpar}
+\caption{\small Normalization semantics for the type-level language.
+Missing rules (including error propagation rules and normalization of quoted internal terms and types) are unsurprising and will be given later.}
+\label{tleval}
+\end{figure*}
+\begin{figure}[t]
 \small\begin{flalign}
@@
 }\}
-\}
+\};\\
+& \F{def}\tvar{concat} = \tinl{\kunit, \FF{R}+\FF{R}}{\tunit};\\
+& \F{def}\tvar{subst} = \tlam{r}{\FF{R}}{\tinr{\kunit, \FF{R}+\FF{R}}{\tinl{\FF{R},\FF{R}}{\tvar{r}}}};\\
+& \F{def}\tvar{coerce} = \tlam{r}{\FF{R}}{\tinr{\kunit, \FF{R}+\FF{R}}{\tinr{\FF{R},\FF{R}}{\tvar{r}}}}
 \end{flalign}
-\caption{Definition of ...}
+\caption{Definition of $\phi_S$, which enables the embedding of fragment $S$ into $\lamAce$.}
 \end{figure}
-\begin{figure}
-\[
-\begin{array}{lcl}
-% & & Definition & Kind\\
-\tvar{concat} & := & \tinl{\kunit, R+R}{b}\\%& \kunit + (R + R)\\
-\tvar{replace} & := & \tlam{r}{R}{\tinr{\kunit, R+R}{\tinl{R,R}{\tvar{r}}}}\\% & \karrow{a}{b}\\
-\tvar{coerce} & := & \tlam{r}{R}{\tinr{\kunit, R+R}{\tinr{R,R}{\tvar{r}}}}\\
-\end{array}
-\]
-\caption{Definitions}
-\end{figure}
+%\begin{figure}
+%\[
+%\begin{array}{lcl}
+%% & & Definition & Kind\\
+%\tvar{concat} & := & \tinl{\kunit, \FF{R}+\FF{R}}{\tunit}\\%& \kunit + (R + R)\\
+%\tvar{replace} & := & \tlam{r}{\FF{R}}{\tinr{\kunit, \FF{R}+\FF{R}}{\tinl{\FF{R},\FF{R}}{\tvar{r}}}}\\% & \karrow{a}{b}\\
+%\tvar{coerce} & := & \tlam{r}{\FF{R}}{\tinr{\kunit, \FF{R}+\FF{R}}{\tinr{\FF{R},\FF{R}}{\tvar{r}}}}\\
+%\end{array}
+%\]
+%\caption{Definitions}
+%\end{figure}
 

### 2014-04-16 sanitation-plas14/sanitation-plas14.tex
lunch edits of PLAS paper.

--- 2014-04-16 sanitation-plas14/sanitation-plas14.tex
@@
 %%
+\newcommand{\lambdas}{\lambda_S}
+\newcommand{\lambdap}{\lambda_P}
+
 % Source language terms.
-\newcommand{\sisubst}[3]{{\tt subst}[#1](#2,#3)}
-\newcommand{\coerce}[2]{ {\tt coerce}[#1](#2)}
-\newcommand{\sistr}[1]{{\tt si\_str}[#1]}
+\newcommand{\sisubst}[3]{{\sf rreplace}[#1](#2,#3)}
+\newcommand{\rssreplace}[3]{\sisubst{#1}{#2}{#2}} % TODO-nrf fix this.
+
+\newcommand{\coerce}[2]{ {\sf rcoerce}[#1](#2)}
+\newcommand{\sistr}[1]{{\sf rstr}[#1]}
 \newcommand{\strin}[1]{\sistr{#1}}
-\newcommand{\siconcat}[2]{{\tt siconcat}(#1,#2)}
+\newcommand{\rsconcat}[2]{{\sf rconcat}(#1,#2)}
 
 % Source language types.
-\newcommand{\stringin}[1]{{\tt string\_in}[#1]}
+\newcommand{\stringin}[1]{{\sf stringin}[#1]}
 
 % target language terms.
-\newcommand{\tsubst}[3]{{\tt tsubst}(#1,#2,#3)}
-\newcommand{\tcheck}[2]{{\tt tcheck}(#1, #2)}
-\renewcommand{\tstr}[1]{{{\tt tstr}[#1]}}
-\newcommand{\tconcat}[2]{{\tt tconcat}(#1,#2)}
+\newcommand{\tsubst}[3]{{\sf replace}(#1,#2,#3)} % TODO-nrf rename the commands.
+\newcommand{\tcheck}[2]{{\sf check}(#1, #2)}
+\renewcommand{\tstr}[1]{{{\sf str}[#1]}}
+\newcommand{\tconcat}[2]{{\sf concat}(#1,#2)}
+\newcommand{\regext}[1]{ {\sf rx}[#1] } % TODO-nrf remove
+\newcommand{\rx}[1]{ {\sf rx}[#1] }
 
 % Target language types
-\newcommand{\str}{{\tt string}}
-\newcommand{\regext}[1]{ {\tt rx}[#1] }
-\newcommand{\rx}[1]{ {\tt rx}[#1] }
+\newcommand{\str}{{\sf string}}
+\newcommand{\regex}{{\sf regex}}
 
 % Meta-theoretic functions
-\newcommand{\dosubst}[3]{{\tt dosubst}(#1,#2,#3)}
-\newcommand{\lsubst}[3]{{\tt lsubst}(#1,#2,#3)}
+\newcommand{\dosubst}[3]{{\tt replace}(#1,#2,#3)}
+\newcommand{\lsubst}[3]{{\tt lreplace}(#1,#2,#3)}
 
 % Judgements
-\newcommand{\err}{\ {\tt err}}
+\newcommand{\err}{\ {\sf err}}
 \newcommand{\trden}[1]{\llbracket #1 \rrbracket} % = Translation Denotation.
@@
 \begin{grammar}
-<r> ::= $\epsilon$ | $.$ | $a$ | $r \cdot r$ | $r + r$ | $r*$ \hfill Regex ($a \in \Sigma$).
+<r> ::= $\epsilon$ | $.$ | $a$ | $r \cdot r$ | $r + r$ | $r*$ \hfill $a \in \Sigma$
 
-<$\psi$> ::=				\hfill	Source language types					\\
-$\stringin{r}$				\hfill Regular expression types \alt
+\caption{Regular expressions over the alphabet $\Sigma$.}
+\end{grammar}
+\end{figure}
 
-<S> ::= \hfill Source language expressions \\
-      $\strin{S}$ \alt
-      $\siconcat{S}{S}$ \hfill String concatenation \alt
+\begin{figure}
+\begin{grammar}
+
+<$\psi$> ::=	...			\hfill	source types					\alt
+$\stringin{r}$
+
+<S> ::= ... \hfill source terms \alt
+      $\strin{s}$ \hfill $s \in \Sigma^{*}$ \alt
+      $\rsconcat{S}{S}$ \alt
       $\sisubst{r}{S}{S}$ \alt
       $\coerce{r}{S}$
+\caption{Syntax for the string sanitation fragment of our source language, $\lambdas$.}
+\end{grammar}
+\end{figure}
 
-<$\theta$> ::= \hfill Target language types \\
-$\str$ \alt $\regext$
+\begin{figure}
+\begin{grammar}
 
-<P> ::= \hfill Internal expressions \\
-  $\tstr{P}$ \alt
-  $\siconcat{P}{P}$ \alt
-  $\tsubst{r}{P}{P}$ \alt
-  $\tcheck{r}{P}$
+<$\theta$> ::= ... \hfill target types \alt
+$\str$ \alt $\regex$
+
+<P> ::= ... \hfill target terms \alt
+  $\tstr{s}$ \alt
+  $\rx{r}$ \alt
+  $\tconcat{P}{P}$ \alt
+  $\tsubst{P}{P}{P}$ \alt
+  $\tcheck{P}{P}$ 
   \end{grammar}
-\caption{Syntax for the string sanitation fragment of $\lcs$}
+\caption{Syntax for the fragment of our target language, $\lambdap$, containing strings and statically constructed regular expressions.}
 \label{fig:lcsSyntax}
@@
 
-\newcommand{\sctx}{\Gamma} % Context for external typing
-\newcommand{\tctx}{\Omega} % Context for internal typing
+\newcommand{\sctx}{\Psi} % Context for external typing
+\newcommand{\tctx}{\Theta} % Context for internal typing
 
@@
 \begin{mathpar}
-
-\inferrule[S-T-include]
+\inferrule[S-T-Stringin-I]
 {s \in \lang{r}}
@@
 {\sctx \vdash S_1 : \stringin{r_1} \\ \sctx \vdash S_2 : \stringin{r_2}}
-{\sctx \vdash \siconcat{S_1}{S_2} : \stringin{r_1 \cdot r_2}}
+{\sctx \vdash \rsconcat{S_1}{S_2} : \stringin{r_1 \cdot r_2}}
 
-\inferrule[S-T-subst]
-{\sctx \vdash S_1 : \stringin{r_1} \\ \sctx \vdash S_2 : \stringin{r_2} \\ {\tt lsubst}(r,r_1,r_2) = r'}
+\inferrule[S-T-replace]
+{\sctx \vdash S_1 : \stringin{r_1} \\ \sctx \vdash S_2 : \stringin{r_2} \\\\ {\tt lsubst}(r,r_1,r_2) = r'}
 {\sctx \vdash \sisubst{r}{S_1}{S_2} : \stringin{r'}}
@@
 \inferrule[S-T-coerce]
-{\sctx \vdash S : \stringin{r}}
-{\sctx \vdash \coerce{r'}{s} : \stringin{r'}}
+{\sctx \vdash S : \stringin{r'}}
+{\sctx \vdash \coerce{r}{S} : \stringin{r}}
 
@@
 
-\caption{Typing rules for source langauge terms (S). The metavariable s ranges over
-string literals.}
+\caption{Typing rules for our fragment of $\lambdas$. 
+The typing context $\sctx$ is standard.}
 \label{fig:etyping}
@@
 $\fbox{\inferrule{}{S \sreduces S}}$
+$\fbox{\inferrule{}{S \err}}$
 \begin{mathpar}
-
-\inferrule[S-E-Include]
+\inferrule[S-E-RStr]
 { \ }
@@
 {S_1 \sreduces \strin{s_1} \\ S_2 \sreduces \strin{s_2}} 
-{\siconcat{S_1}{S_2} \sreduces \strin{s_1 \cdot s_2}} % ???
+{\rsconcat{S_1}{S_2} \sreduces \strin{s_1 s_2}} % ???
 
-\inferrule[S-E-Subst]
+\inferrule[S-E-Replace]
 {S_1 \sreduces \strin{s_1} \\ S_2 \sreduces \strin{s_2} \\ \dosubst{r}{s_1}{s_2} = s} 
-{\sisubst{r}{S_1}{S_2} \sreduces s}
+{\sisubst{r}{S_1}{S_2} \sreduces \sistr{s}}
 
@@
 
-\inferrule[S-E-Coerce-NotOk]
-{\coerce{r}{S} \sreduces \strin{s} \\ s \not \in \lang{r}}
-{S \err}
+\inferrule[S-E-Coerce-Err]
+{S \sreduces \strin{s} \\ s \not \in \lang{r}}
+{\coerce{r}{S} \err}
 
 \end{mathpar}
-\caption{Big step semantics for the string fragment of the source language.
-There are also unmentioned error propagation rules for for S-E-Concat and S-E-Subst.}
+\caption{Big step semantics for our fragment of $\lambdas$.
+Error propagation rules are omitted.}
 \label{fig:eeval}
@@
 \begin{mathpar}
-
 \inferrule[P-T-String]
@@
 {\tctx \vdash \tstr{s} : \str}
+
+\inferrule[P-T-Regex]
+{ \ }
+{\tctx \vdash \rx{r} : \regex}
 
@@
 
-\inferrule[P-T-Subst]
-{\tctx \vdash P_1 : \str \\ \tctx \vdash P_2 : \str}
-{\tctx \vdash \tsubst{\rx{r}}{P_1}{P_2} : \str}
+\inferrule[P-T-Replace]
+{\tctx \vdash P_1 : \regex \\ \tctx \vdash P_2 : \str \\ \tctx \vdash P_3 : \str }
+{\tctx \vdash \tsubst{P_1}{P_2}{P_3} : \str}
 
 \inferrule[P-T-Check]
-{\tctx \vdash P : \str}
-{\tctx \vdash \tcheck{\rx{r}}{s} : \str}
+{\tctx \vdash P_1 : \regex \\ \tctx \vdash P_2 : \str \\ }
+{\tctx \vdash \tcheck{P_1}{P_2} : \str}
 
 \end{mathpar}
-\caption{Typing rules for target langauge terms (P). The metavariable s ranges over
-string literals.}
+\caption{Typing rules for our fragment of $\lambdap$.
+The typing context $\tctx$ is standard.}
 \label{fig:etyping}
@@
 $\fbox{\inferrule{}{P \treduces P}}$
+$\fbox{\inferrule{}{P \err}}$
+
 \begin{mathpar}
-\inferrule[P-E-Strin]
+\inferrule[P-E-Str]
+{ \ }
+{\tstr{s} \treduces \tstr{s}}
+
+\inferrule[P-E-Rx]
 { \ }
@@
 
-\\
-
 \inferrule[P-E-Concat]
 {P_1 \treduces \tstr{s_1} \\ P_2 \treduces \tstr{s_2}} 
-{\tconcat{P_1}{P_2} \treduces \tstr{s_1 \cdot s_2}} % ???
+{\tconcat{P_1}{P_2} \treduces \tstr{s_1 s_2}} % ???
 
-\inferrule[P-E-Subst]
+\inferrule[P-E-Replace]
 {P_1 \treduces \rx{r} \\ P_2 \treduces \tstr{s_2} \\ P_3 \treduces \tstr{s_3} \\ \dosubst{r}{s_2}{s_3} = s} 
-{\tsubst{r}{P_2}{P_3} \treduces \tstr{s}}
+{\tsubst{P_1}{P_2}{P_3} \treduces \tstr{s}}
 
 \inferrule[P-E-Check-OK]
-{P \treduces \strin{s} \\ s \in \lang{r}}
-{\tcheck{r}{P} \treduces \tstr{s}}
+{P_1 \treduces \rx{r} \\ P_2 \treduces \strin{s} \\ s \in \lang{r}}
+{\tcheck{P_1}{P_2} \treduces \tstr{s}}
 
-\inferrule[P-E-Check-NotOk]
-{\tcheck{r}{P} \treduces \tstr{s} \\ s \not \in \lang{r}}
-{P \err}
+\inferrule[P-E-Check-Err]
+{P_1 \treduces \rx{r} \\ P_2 \treduces \tstr{s} \\ s \not \in \lang{r}}
+{\tcheck{P_1}{P_2} \err}
 \end{mathpar}
-\caption{Big step semantics for the target lanugage.
-There are also unimplemented error propagation rules for for P-E-Concat and P-E-Subst.}
+\caption{Big step semantics for our fragment of $\lambdap$.
+Error propagation rules are omitted.}
 \label{fig:ieval}
@@
 { \trden{S_1} = P_1 \\ \trden{S_2} = P_2}
-{\trden{\siconcat{S_1}{S_2}} = \tconcat{P_1}{P_2}}
+{\trden{\rsconcat{S_1}{S_2}} = \tconcat{P_1}{P_2}}
 
@@
 Similarly for $T_2$ and $S_2$.
-In the error cases, the propagations rules corresponding to S-E-Concat and T-E-Concat imply that $\tconcat{T_1}{T_2} \err$ and $\siconcat{S_1}{S_2} \err$.
-In the non-error case, $\siconcat{S_1}{S_2} \sreduces \strin{s_1 \cdot s_2}$ by S-E-Concat and $\tconcat{T_1}{T_2} \treduces \tstr{s_1 \cdot s_2}$ by T-E-Concat.
-Finally, $s_1 \in \lang{r_1}$ and $s_2 \in \lang{r_2}$ implies that $s_1 \cdot s_2 \in \lang{r}$ by \ref{lemmaregex}.
+In the error cases, the propagations rules corresponding to S-E-Concat and T-E-Concat imply that $\tconcat{T_1}{T_2} \err$ and $\rsconcat{S_1}{S_2} \err$.
+In the non-error case, $\rsconcat{S_1}{S_2} \sreduces \strin{s_1 s_2}$ by S-E-Concat and $\tconcat{T_1}{T_2} \treduces \tstr{s_1 s_2}$ by T-E-Concat.
+Finally, $s_1 \in \lang{r_1}$ and $s_2 \in \lang{r_2}$ implies that $s_1 s_2 \in \lang{r}$ by \ref{lemmaregex}.
 \item[Tr-subst]

### 2014-05-10 ace-oopsla14/ace-oopsla14.tex
fixed some things

@@
 \usepackage{amsthm}
+\newtheorem{theorem}{Theorem}
+\newtheorem{lemma}{Lemma}
+\newtheorem{corollary}{Corollary}
 \usepackage{ stmaryrd }
@@
 
-def print_transfer(t):
+def log_transfer(t):
   print ("Transferring %s to %s." % (
@@
 
-\section{Theoretical Foundations}\label{theory}
-\newcommand{\F}[1]{\textsf{#1}~}
-\newcommand{\FF}[1]{\textsf{#1}}
+\section{@$\lambda$: Foundations of Active Typechecking and Translation}\label{theory}
+\newcommand{\F}[1]{{\sf #1}~}
+\newcommand{\FF}[1]{{\sf #1}}
 \newcommand{\Q}{\FF{Arg}}
 \renewcommand{\tnil}[1]{[]}
-\begin{figure}[t]
+\newcommand{\ltxt}[1]{\ell_{\text{#1}}}
+\begin{figure*}[t]
 \small
-$$\begin{array}{rcl}	
-%\textbf{programs} & 
-\rho & ::= & \F{tycon}\fvar{tycon}~\F{of}\kappaidx~{\{}TC{\}}; \rho \pipe e\\ %\pfam{\familyDf}{\progsort}
-%\\&  \pipe & \pdef{t}{\kappa}{\tau}{\progsort}  \pipe e
-%\pipe \pdef{t}{\kappa}{\tau}{\progsort} 
- TC & ::= & \F{iana} \{\tau\}; \F{isyn} \{\tau\}; \F{esyn} \{\tau\}; \F{rep} \{\tau\}\\
-\\ 
-e & ::= & \evar{x} \pipe \elam{\evar{x}}{\tau}{e} \pipe I \pipe I : \tau \pipe I :: \fvar{tycon} \pipe E\\
-I & ::= &  \FF{intro}[\tauidx](\splat{e}{1}{n})\\
-E & ::= & e\cdot\FF{elim}[\tauidx](\splat{e}{1}{n})\\
-\\
- \tau 	& ::= 	& 	\tvar{t} \pipe 
-														\tlam{t}{\kappa}{\tau} \pipe 
-														\tapp{\tau_1}{\tau_2}  
-\\&\pipe&											
-														\tnil{\kappa} \pipe \tcons{\tau_1}{\tau_2} \pipe 
-									                     \tfold{\tau_1}{\tau_2}{h}{t}{r}{\tau_3}
-														\\
- 		& \pipe	& 	  \ell \pipe 
-		\tunit \pipe 
-														\tpair{\tau_{1}}{\tau_{2}} \pipe 
-														\tfst{\tau} \pipe 
-														\tsnd{\tau} \pipe \cdots 
-														\\	
-%     & \pipe & \tinl{\kappa_2}{\tau_1} \pipe \tinr{\kappa_1}{\tau_2} \\
-%     &\pipe& \tsumcase{\tau}{t}{\tau_1}{t}{\tau_2} \\
-&\pipe	& 	\ttype{tycon}{\tauidx} \\&\pipe& \tfamcase{\tau}{tycon}{x}{\tau_1}{\tau_2}\\
-&\pipe & 					\tifeq{\tau_{1}}{\tau_{2}}{\kappa}{\tau_{3}}{\tau_{4}} 
-%														\\													%				& & \pipe & \tfamcase{\tau}{Fam}{x}{\tau_1}{\tau_2}\\
-%																								
-\\ 		 & 	\pipe	&	\tden{\tau_2}{\tau_1} \pipe \terr %\pipe \tden{\ibar}{\tau}^{\checkmark} 
-\pipe \ttypeof{\tau} \pipe \itransof{\tau} \\% \tdencase{\tau}{x}{t}{\tau_1}{\tau_2}\\
-% %& & \pipe & 
-%%														\tdencase{\tau}{y}{x}{\tau_1}{\tau_2}
-%%														 \\
-%
- & \pipe & \titerm{\iota}
-			\pipe	\titype{\sigma} \pipe \trepof{\tau} \\
-		& \pipe &  (\Gamma; e)? \pipe \FF{syn}(\tau) \pipe \FF{ana}(\tau; \tau')\\
-%\textbf{external terms} 				&	e	&	::=	&	\evar{x} \pipe 
-%%														\efix{x}{\tau}{e} \pipe 
-%														\elam{\evar{x}}{\tau}{e} \pipe 
-%														\eop{tycon}{op}{
-%															\tauidx
-%														}{
-%  												    		\splat{e}{1}{n}
-%														} \\
-%									& 		&		& 	\\
-%
-%
-%												
-%%\text{deabstracted}& \iota & ::= & \mathcal{G}[\iota, \sigma]\\
-%												
-%\\
-%\\
-%							
-%% &  & \pipe & \tvalof{\tau_1}{\tau_2} \pipe \iup{\tau} \\
-%% 												\trepof{\tau} \pipe \dup{\tau}\\
-%\text{translational IL}	& \bar{\iota} & ::= & x \pipe \ifix{x}{\bar \sigma}{\bar \iota} 
-%	%\pipe \ilam{x}{\bar \sigma}{\bar \iota} \pipe \iapp{\bar \iota_1}{\bar \iota_2} 
-%	\pipe \cdots \pipe \itransof{\tau} \\						
-% & \bar{\sigma} & ::= & \darrow{\bar \sigma_1}{\bar \sigma_2} \pipe \cdots \pipe \dup{\tau} \pipe \trepof{\tau} \\
-%%\text{abstracted} & \sabs & ::= & \darrow{\sabs_1}{\sabs_2} \pipe \cdots \pipe \sabsrep{\tau}
-%											\\
-\kappa	&	::=	&	\karrow{\kappa_1}{\kappa_2} \pipe \klist{\kappa} \pipe \FF{L} \pipe \kunit \pipe \kpair{\kappa_1}{\kappa_2} \pipe \cdots \\
- & \pipe & \kTypeBlur \pipe \FF{TT} \pipe \kITerm \pipe \kIType \pipe \Q
-%											    \klabel \pipe
-%											    \klist{\kappa} \pipe
-%												\kunit \pipe 
-%												\kpair{\kappa_{1}}{\kappa_{2}} \\
-%												&&\pipe&
-%												\ksum{\kappa_1}{\kappa_2} \pipe
-%												\kTypeBlur \pipe \kDen \pipe 
-%												\kIType								
-%%\textbf{ops signature}			& \Theta	&	::=	&	\kOpEmpty \pipe \kOp{\Theta}{op}{\kappai}\\
-%%											 							&		&		&	\\
-\\
-\\
-\iota	&	::=	&	\evar{x} \pipe 
-												\ifix{\evar{x}}{\sigma}{\iota} \pipe
-												\ilam{\evar{x}}{\sigma}{\iota} \pipe 
-												\iapp{\iota_{1}}{\iota_{2}} 
-												\\ & \pipe & \iunit \pipe \ipair{\iota_1}{\iota_2} \pipe \ifst{\iota} \pipe \isnd{\iota} \pipe \cdots \pipe \iup{\tau}
-%\\&\pipe&								\cdots \pipe 	
-%												n \pipe \iop{\iota_{1}}{\iota_{2}} \pipe \iIfEq{\iota_{1}}{\iota_{2}}{\dint}{\iota_{3}}{\iota_{4}} \\
-%& \pipe & 
-%												\iunit \pipe
-%												\ipair{\iota_{1}}{\iota_{2}} \pipe 
-%												\ifst{\iota} \pipe
-%												\isnd{\iota} 
-%\\&\pipe&												
-%												 \iinl{\sigma_2}{\iota_1} \pipe \iinr{\sigma_1}{\iota_2} \\&\pipe & \icase{e}{x}{e_1}{x}{e_2} \\
-\\	\sigma	&	::=	&    \darrow{\sigma_1}{\sigma_2} \pipe \dunit \pipe \dpair{\sigma_1}{\sigma_2} %\pipe \dint \pipe \dunit \pipe \dpair{\sigma_1}{\sigma_2} \pipe \dsum{\sigma_1}{\sigma_2}  
-\pipe \cdots \pipe \dup{\tau}
-\end{array}$$
+$
+\begin{array}{l}
+\F{using} \phi_{\texttt{nat}}; \phi_{\texttt{lprod}}\\
+\elet{one}{\tvar{s}\langle\tvar{z}\langle\rangle\rangle}{\\
+\elet{plus}{(\lambda x.\lambda y.\tvar{natrec}~x~\langle y; \lambda p.\lambda r.\tvar{s}\langle r\rangle\rangle) : \tvar{nat} \rightharpoonup \tvar{nat} \rightharpoonup \tvar{nat}}{\\
+\FF{let}~y = \{\ltxt{1}=one, \ltxt{2}=plus~one~one\} : \ttype{lprod}{(\ltxt{1}, \tvar{nat}) :: (\ltxt{2}, \tvar{nat}) :: []}}}~\FF{in}\\
+y.\ell_2
+\end{array}
+$
+\caption{\small A program that uses the natural number fragment, $\phi_\texttt{nat}$, defined in Figure \ref{nat-atfrag}, written with the syntactic sugar defined in Figure \ref{desugaring}.}
+\label{nat-sugared}
+\end{figure*}
+\begin{figure*}[t]
+\small
+\hspace{-5px}
+$
+\begin{array}{l}
+\F{using} \hat\phi_{\texttt{nat}}; \hat\phi_{\texttt{lprod}}\\
+\elet{one}{\FF{intro}[\FF{in}[\ltxt{s}](\tunit)](\FF{intro}[\FF{in}[\ltxt{z}](\tunit)]()) : \ttype{nat}{\tunit}}{\\
+\elet{plus}{(\lambda x.\lambda y.x\cdot\FF{elim}[()](y; \lambda p.\lambda r.\FF{intro}[\FF{in}[\ltxt{s}]()](r)) :\ttype{parr}{(\ttype{nat}{\tunit}, \ttype{parr}{(\ttype{nat}{\tunit}, \ttype{nat}{\tunit})})}}{\\
+\FF{let}~y = \FF{intro}[\ell_1 :: \ell_2 :: []](one; plus\cdot\FF{elim}[()](one)\cdot\FF{elim}[()](one)) : \ttype{lprod}{(\ltxt{1}, \tvar{nat}) :: (\ltxt{2}, \tvar{nat}) :: []}}}~\FF{in}\\
+y\cdot\FF{elim}[\ell_2]()
+\end{array}
+$
+\caption{\small An equivalent program written in core @$\lambda$ without syntactic sugar and with type-level terms normalized.}
+\label{nat-desugared}
+\end{figure*}
+\begin{figure*}
+\small
+$
+\begin{array}{lcl}
+\phi_{\texttt{nat}} & := & \F{tycon}\fvar{nat}~\F{of}\kunit~\{\delta_{\texttt{nat}}\}\\
+&& \F{def}\tvar{nat}:\kTypeBlur = \ttype{nat}{\tunit};\\
+&& \F{def}\tvar{z}:\Pi[\{\ltxt{idx}\hookrightarrow\kappa_{\text{nat-intro-idx}},\ltxt{ty}\hookrightarrow\kTypeBlur\}]={(\ltxt{idx}=\FF{in}[\ltxt{z}](\tunit), \ltxt{ty}=\tvar{nat})};\\
+&& \F{def}\tvar{s}:\Pi[\{\ltxt{idx}\hookrightarrow\kappa_{\text{nat-intro-idx}},\ltxt{ty}\hookrightarrow\kTypeBlur\}]={(\ltxt{idx}=\FF{in}[\ltxt{s}](\tunit), \ltxt{ty}=\tvar{nat})};\\
+&& \F{def}\tvar{natrec}:\kappa_{\text{nat-elim-idx}} =\tunit\\
+\kappa_{\text{nat-intro-idx}} & := & \Sigma[\{l_z \hookrightarrow \kunit, l_s \hookrightarrow \kunit\}]\\
+\kappa_{\text{nat-elim-idx}} & := & \kunit\\
+\end{array}
+$
 %\vspace{-10pt}
-\caption{\small Syntax of Core $\lamAce$.
-Here, $x$ ranges over external and internal language variables, $\tvar{t}$ ranges over type-level variables, $\fvar{tycon}$ ranges over type constructor names and $\ell$ ranges over labels.
-\label{grammar}}
-\end{figure}
-\begin{figure}
-\small
-\begin{flalign}
-& \F{tycon}\fvar{record}~\F{of}\klist{\kpair{\FF{L}}{\kTypeBlur}}~\{\\
-& \quad \FF{iana}~\{\tlam{i}{\kpair{\klist{\kpair{\FF{L}}{\kTypeBlur}}}{\klist{\FF{L}}}}{\tlam{a}{\klist{\Q}}{\\& \quad\quad\tfold{\tvar{zip3}~\tfst{\tvar{i}}~\tsnd{\tvar{i}}~\tvar{a}}
-	 {\titerm{\tunit}}{h}{t}{r}{
-	\\&\quad\quad\quad\tifeq{\tfst{\tvar{first}~\tvar{h}}}{\tvar{second}~\tvar{h}}{\FF{L}}{
-	\\&\quad\quad\quad\quad
-	\F{let}~\tvar{x}:\kITerm~=~\FF{ana}(\tvar{third}~\tvar{h}; \tsnd{\tvar{first}~\tvar{h}})
-~\FF{in}
-\\ & \quad\quad\quad\quad \tfold{\tvar{t}}{\tvar{x}
-}{\_}{\_}{\_}{\titerm{(\iup{\tvar{x}},\iup{\tvar{r}})}}	}{\terr}
-%     \F{let}~\tvar{tt}:\FF{TT}=~\FF{syn}(\tsnd{\tsnd{\tvar{h}}}) \F{in}\\
-%\\     & \quad\quad\quad 
-}
-}}\}\\
-& \quad \FF{isyn}~\{\tlam{i}{\klist{\FF{L}}}{\tlam{a}{\klist{\Q}}{
-\\&\quad\quad \tfold{\tvar{zip2}~\tvar{i}~\tvar{a}}{\tden{\titerm{()}}{\ttype{record}{\tnil{{\kpair{\FF{L}}{\kTypeBlur}}}}}}{h}{t}{r}{
-\\&\quad\quad\quad \F{let}~\tvar{htt}:\FF{TT}~=~\FF{syn}(\tsnd{\tvar{h}})
-~\FF{in}
-\\&\quad\quad\quad \F{let}~\tvar{hty}:\kTypeBlur~=~\ttypeof{\tvar{htt}}
-\\&\quad\quad\quad \F{let}~\tvar{ty}:\kTypeBlur~=~\ttype{record}{(\tfst{\tvar{h}}, \tvar{hty})::\tnil{{\kpair{\FF{L}}{\kTypeBlur}}}}
-~\FF{in}
-\\&\quad\quad\quad \tfold{\tvar t}{\tden{\itransof{\tvar{htt}}}{\tvar{ty}}}{\_}{\_}{\_}{
-\\&\quad\quad\quad\quad \tden{\titerm{(\iup{\itransof{\tvar{htt}}},\iup{\itransof{\tvar{r}}})}}{\tvar{ty}}}}
-}}\}\\
-& \quad \FF{esyn}~\{\tlam{i}{\FF{L}}{\tlam{a}{\klist{\Q}}{
-	\tvar{arity1}~\tvar{a}~\tlam{ty}{\kTypeBlur}{\tlam{x}{\kITerm}{
-\\&\quad\quad		\tfamcase{\tvar{ty}}{record}{sig}{
-\\&\quad\quad\quad \tfold{\tvar{sig}}{\terr}{h}{t}{r}{
-\\&\quad\quad\quad\quad \tifeq{\tfst{\tvar{h}}}{\tvar{i}}{\FF{L}}{\tfold{\tvar{t}}{\tden{\tvar{x}}{\tsnd{\tvar{h}}}}{\_}{t'}{\_}{
-\\&\quad\quad\quad\quad\quad \tfold{\tvar{t'}}{\tden{\titerm{\ifst{\iup{\tvar{x}}}}}{\tsnd{\tvar{h}}}}{\_}{\_}{\tvar{r'}}{
-\\&\quad\quad\quad\quad\quad\quad \tden{\titerm{\isnd{\iup{{\itransof{\tvar r}}}}}}{\tsnd{\tvar{h}}}}}\\&\quad\quad\quad\quad\hspace{-3px}}{\tvar{r}}
-}\\&\quad\quad\hspace{-3px}}{\terr}
-	}}
-}}\\
-& \quad \FF{rep}~\{\tlam{i}{\klist{\kpair{\FF{L}}{\kTypeBlur}}}{ 	\\& \quad \quad \tfold{\tvar{i}}{\titype{\dunit}}{s}{j}{r}{
- 		\tfold{\tvar{j}}{\trepof{\tsnd{\tvar{s}}}}{\_}{\_}{\_}{\\
-		&\quad\quad\quad
- 		\titype{\dpair{\dup{\trepof{\tsnd{\tvar{s}}}}}{\dup{\tvar{r}}}}
- 		}
- 	}}\}\\
-& \}; \\
-& \F{let}\tvar{R} : \kTypeBlur = \ttype{record}{(\ell_1, \ttype{record}{\tnil{{\FF{L}\times\kTypeBlur}}}) :: \tnil{{\FF{L}\times\kTypeBlur}}}~\F{in}\\
-&\F{let}id = \elam{x}{\tvar{R}}{\{\ell_1=~x\cdot\ell_1\} :: \fvar{record}}\\
-&\F{let}triv = id[\{\ell_1 = \{\}\}]\cdot\ell_1
-%&\F{let}one = \FF{intro}[\ell_\text{succ}]
-%&\F{let}plus = \elam{x}{\tvar{nat}}{\elam{y}{\tvar{nat}}{\FF{elim}[()](x; y; \\
-%& \quad \elam{r}{\tvar{nat}}{intro[\ell_\text{succ}](r) : \tvar{nat}}})}
-\end{flalign}
-\caption{The definition of the record type in $\lamAce$ using the desugarings in Figure \ref{desugaring} and with the addition of simple let bindings for both type and term variables (not shown).
-Some type-level helper functions also omitted for concision.}
-
-\label{record-theory}
-\end{figure}
-We will now give a core typed lambda calculus, $\lamAce$, that captures the semantics described in the previous sections.
+\caption{\small The natural number fragment, including definitions used by the assisted intro and elim desugarings, defined and shown being used above.}
+\label{nat-atfrag}
+\end{figure*}
+In this section, we will describe the foundations of \emph{active typechecking and translation} by constructing a core lambda calculus, @$\lambda$. 
+A \emph{program} in @$\lambda$, $\rho$, consists of a series of fragment definitions, $\phi$, for use by an external term, $e$.
+An example of an @$\lambda$ program, $\rho_{\text{example}}$, that uses {fragment definitions} $\phi_{\texttt{nat}}$ and $\phi_{\texttt{lprod}}$, defining primitive natural numbers and labeled products (i.e.
+ordered records), is shown in Fig.
+\ref{nat-sugared}.
+In Ace,  fragment definitions are packaged separately; we do not include this in the core calculus for simplicity.
+
+In this example, we see the introduction and elimination forms for natural numbers, functions and labeled tuples all being used.
+However, in the core syntax, shown (from the perspective of fragment clients) in Fig.
+\ref{client-syntax}, there are only two external introductory forms, $\lambda x.e$ and $\FF{intro}[\sigma](\overline{e})$, and one external elimination form, $e\cdot\FF{elim}[\sigma](\overline{e})$, where $\overline{e}$ is shorthand for zero or more semicolon-separated external terms, called the \emph{arguments} of the introduction or elimination operation.
+The forms used in Fig.
+\ref{nat-sugared} are \emph{derived forms}, defined in terms of these core forms in Fig.
+\ref{derived}.
+
+Both fragments and external terms can contain \emph{static terms}, $\sigma$, which are evaluated statically to \emph{static values}, $\hat\sigma$.
+The static language is itself a simply-typed lambda calculus.
+We call the types of static terms \emph{kinds}, $\kappa$, by convention.
+Static variables are written in bold font, $\tvar{x}$, to emphasize that they are distinct from variables bound by external terms, written $x$.
+For example, in the example in Figure \ref{nat-sugared}, $\tvar{nat}$ is a static variable bound by the fragment $\phi_{\texttt{nat}}$ to the natural number type.
+Types have kind $\kTypeBlur$ and are introduced by naming a \emph{type constructor}, written in small-caps, e.g.
+$\fvar{nat}$ and $\fvar{lprod}$, and providing an \emph{index}, which is a static term of the \emph{index kind} of the type constructor.
+For example, $\fvar{nat}$ is indexed by $\kunit$ (because there is only one natural number type), so $\tvar{nat}$ is defined to be $\ttype{nat}{\tunit}$, and $\fvar{lprod}$ is indexed by $\klist{\kpair{\FF{L}}{\kTypeBlur}}$, classifying static lists pairing \emph{labels} with types.
+Labels are static values of kind $\FF{L}$, written abstractly using the metavariables containing $\ell$ in our examples.
+
+Before typechecking the external term in a program, the program is kind checked and its static terms are normalized to static values.
+We write static values using the hatted metavariable $\hat \sigma$.
+Other syntactic metavariables in our system also have hatted forms if they can contain static terms.
+The desugared, statically normalized version of the example in Figure \ref{nat-sugared}, $\hat\rho_{\text{example}}$, is shown in Figure \ref{nat-normal}.
+%The syntax of @$\lambda$ from the perspective of a fragment client is given in Fig.
+\ref{atlam-syntax}.
+We will see a few additions relevant only to fragment providers when we discuss defining fragments below.
+
+\subsection{External Terms}
+ a program consists of first \emph{kind checking} its static terms, then normalizing its static terms, then typechecking the external term and,  simultaneously, \emph{translating} it to a term, $\iota$, in the \emph{typed internal language}.
+The dynamic behavior of an external term is determined entirely by its translation to this language.
+Internal types are written $\tau$.
+The internal language is only exposed to fragment providers, not to clients (i.e.
+normal developers).
+
+%We will now give a core typed lambda calculus, @$\lambda$, that captures the semantics described in the previous sections.
 It is intended to make precise how active typechecking and translation works and how our mechanism relates to existing work on bidirectional typechecking, type-level computation and typed compilation while abstracting away from the details of Python's syntax and imposing a stronger type-level semantics that will allow us to state metatheoretic properties of interest.
 We will assume a fixed base for functions (providing the standard semantics of lambda functions) and target language, which we here call the \emph{internal language}.
-
-The syntax of $\lamAce$ is shown in Figure \ref{grammar} and an example \emph{program} that defines a type constructor, $\fvar{record}$, with a semantics similar to that given in Listing \ref{record}\footnote{Technically, this defines labeled tuples, because the order of labels matters.}, using it to write an identity function and compute the value $triv$ (the empty record), is shown in Figure \ref{record-theory}.
+ % program written in core @$\lambda$, without syntactic sugar and with all static terms normalized, is shown in Fig.
+\ref{nat-desugared}.
+The syntax of core @$\lambda$ is shown in Fig.
+\ref{grammar} and the syntactic desugarings are shown in Fig.
+\ref{desugaring}.
+The definition of $\phi_{\texttt{nat}}$ is shown in Fig.
+\ref{nat-atfrag}.
+We will discuss these in the following sections.
 
@@
 
-\subsection{Overview}\label{programs}
-
-A \emph{program}, $\rho$, consists of a series of constructor declarations followed by an external term, $e$.
-Compiling a program consists of first \emph{kind checking} it (see below), then typechecking the external term and simultaneously {translating} it to a term, $\iota$, in the {typed internal language}.
+%\begin{figure}[t]
+%\small
+%$$\begin{array}{rrcl}	
+%%\textbf{programs} & 
+%\text{programs} & \rho & ::= & \F{using}\phi~\F{in}e \\
+%\text{fragments} & \phi &  ::= & \F{tycon}\fvar{tycon}~\F{of}\kappat{tyidx}~{\{}\F{iana} \tau_1; \F{esyn} \tau_2; \F{rep} \tau_3{\}} %\pipe \phi; \phi %\\
+%%\\
+%%& \pipe & 
+%\pipe \F{def}\tvar{t} : \kappa = \tau \pipe \phi; \phi%\pdef{t}{\kappa}{\tau}{\progsort} \pipe 
+%\\ %\pfam{\familyDf}{\progsort}
+%%\\&  \pipe & \pdef{t}{\kappa}{\tau}{\progsort}  \pipe e
+%%\pipe \pdef{t}{\kappa}{\tau}{\progsort} 
+%\text{external terms}& e & ::= & \evar{x} \pipe \F{let} x = e_1~\F{in} e_2 \pipe \lambda x.e \pipe e : \tau\\
+%&& \pipe & \FF{intro}[\taut{opidx}](\splat{e}{1}{n})\pipe e\cdot\FF{elim}[\taut{opidx}](\splat{e}{1}{n}) \\
+%
+%\text{type-level terms} & \tau 	& ::= 	& 	\tvar{t} \pipe \tifeq{\tau_{1}}{\tau_{2}}{\kappa}{\tau_{3}}{\tau_{4}} \pipe \terr  \\
+% && \pipe & 
+% %\F{let}\tvar{t}{:}\kappa=\tau_1~\F{in} \tau_2 \pipe 
+%														\tlam{t}{\kappa}{\tau} \pipe 
+%														\tapp{\tau_1}{\tau_2} \\
+%&&\pipe&											
+%														\tnil{\kappa} \pipe \tcons{\tau_1}{\tau_2} \pipe 
+%									                     \tfold{\tau_1}{\tau_2}{\tvar{t}_{hd}}{\tvar{t}_{tl}}{\tvar{t}_{rec}}{\tau_3}
+%														\\
+% 		&& \pipe	& 	  \ell  \\
+%	    && \pipe &  
+%		\tunit \pipe 
+%														\tpair{\tau_{1}}{\tau_{2}} \pipe 
+%														\tfst{\tau} \pipe 
+%														\tsnd{\tau} 
+%														\\	
+%     && \pipe & \tinl{\kappa_1, \kappa_2}{\tau_1} \pipe \tinr{\kappa_1, \kappa_2}{\tau_2} \pipe \tsumcase{\tau}{t}{\tau_1}{t}{\tau_2}\\
+%&&\pipe	& 	\ttype{tycon}{\taut{tyidx}} \pipe  \tfamcase{\tau}{tycon}{t}{\tau_1}{\tau_2}\\
+%%														\\													%				& & \pipe & \tfamcase{\tau}{Fam}{x}{\tau_1}{\tau_2}\\
+%%																								
+%%\\ 		 & 	\pipe	&	\tden{\tau_2}{\tau_1} \pipe \terr %\pipe \tden{\ibar}{\tau}^{\checkmark} 
+%%\pipe \ttypeof{\tau} \pipe \itransof{\tau} \\% \tdencase{\tau}{x}{t}{\tau_1}{\tau_2}\\
+%% %& & \pipe & 
+%%%														\tdencase{\tau}{y}{x}{\tau_1}{\tau_2}
+%%%														 \\
+%%
+% && \pipe & \titerm{\iota}
+%			\pipe	\titype{\sigma} \pipe \trepof{\tau} \\
+%		&& \pipe &  (\Gamma; e)? \pipe \FF{syn}(\tau_1; \tvar{t}_{ty},\tvar{t}_{trans}.\tau_2) \pipe \FF{ana}(\tau_1; \tau_2; \tvar{t}_{trans}.\tau_3)\\
+%%\textbf{external terms} 				&	e	&	::=	&	\evar{x} \pipe 
+%%%														\efix{x}{\tau}{e} \pipe 
+%%														\elam{\evar{x}}{\tau}{e} \pipe 
+%%														\eop{tycon}{op}{
+%%															\tauidx
+%%														}{
+%%  												    		\splat{e}{1}{n}
+%%														} \\
+%%									& 		&		& 	\\
+%%
+%%
+%%												
+%%%\text{deabstracted}& \iota & ::= & \mathcal{G}[\iota, \sigma]\\
+%%												
+%%\\
+%%\\
+%%							
+%%% &  & \pipe & \tvalof{\tau_1}{\tau_2} \pipe \iup{\tau} \\
+%%% 												\trepof{\tau} \pipe \dup{\tau}\\
+%%\text{translational IL}	& \bar{\iota} & ::= & x \pipe \ifix{x}{\bar \sigma}{\bar \iota} 
+%%	%\pipe \ilam{x}{\bar \sigma}{\bar \iota} \pipe \iapp{\bar \iota_1}{\bar \iota_2} 
+%%	\pipe \cdots \pipe \itransof{\tau} \\						
+%% & \bar{\sigma} & ::= & \darrow{\bar \sigma_1}{\bar \sigma_2} \pipe \cdots \pipe \dup{\tau} \pipe \trepof{\tau} \\
+%%%\text{abstracted} & \sabs & ::= & \darrow{\sabs_1}{\sabs_2} \pipe \cdots \pipe \sabsrep{\tau}
+%%											\\
+%\text{kinds} & \kappa	&	::=	&	\karrow{\kappa_1}{\kappa_2} \pipe \klist{\kappa} \pipe \FF{L} \pipe \kunit \pipe \kpair{\kappa_1}{\kappa_2} \pipe \ksum{\kappa_1}{\kappa_2} \pipe \kTypeBlur \pipe \kITerm \pipe \kIType \pipe \Q
+%%											    \klabel \pipe
+%%											    \klist{\kappa} \pipe
+%%												\kunit \pipe 
+%%												\kpair{\kappa_{1}}{\kappa_{2}} \\
+%%												&&\pipe&
+%%												\ksum{\kappa_1}{\kappa_2} \pipe
+%%												\kTypeBlur \pipe \kDen \pipe 
+%%												\kIType								
+%%%\textbf{ops signature}			& \Theta	&	::=	&	\kOpEmpty \pipe \kOp{\Theta}{op}{\kappai}\\
+%%%											 							&		&		&	\\
+%\\
+%\text{typing contexts} & \Gamma & ::= & \emptyset \pipe \Gamma, x \Rightarrow \tau\\
+%\\
+%\text{internal terms} & \iota	&	::=	&	\evar{x} \pipe 
+%												\ifix{\evar{x}}{\sigma}{\iota} \pipe
+%												\ilam{\evar{x}}{\sigma}{\iota} \pipe 
+%												\iapp{\iota_{1}}{\iota_{2}} \pipe n \pipe \iota_1 \pm \iota_2 \pipe \FF{if0}(\iota; \iota_1; \iota_2)\\
+%		&& \pipe &  \iunit \pipe \ipair{\iota_1}{\iota_2} \pipe \ifst{\iota} \pipe \isnd{\iota} \pipe \iup{\tau} 
+%%\\&\pipe&								\cdots \pipe 	
+%%												n \pipe \iop{\iota_{1}}{\iota_{2}} \pipe \iIfEq{\iota_{1}}{\iota_{2}}{\dint}{\iota_{3}}{\iota_{4}} \\
+%%& \pipe & 
+%%												\iunit \pipe
+%%												\ipair{\iota_{1}}{\iota_{2}} \pipe 
+%%												\ifst{\iota} \pipe
+%%												\isnd{\iota} 
+%%\\&\pipe&												
+%%												 \iinl{\sigma_2}{\iota_1} \pipe \iinr{\sigma_1}{\iota_2} \\&\pipe & \icase{e}{x}{e_1}{x}{e_2} \\
+%\\	\text{internal types} & \sigma	&	::=	&    \darrow{\sigma_1}{\sigma_2} \pipe \dint \pipe \dunit \pipe \dpair{\sigma_1}{\sigma_2} %\pipe \dint \pipe \dunit \pipe \dpair{\sigma_1}{\sigma_2} \pipe \dsum{\sigma_1}{\sigma_2}
+% \pipe \dup{\tau}
+%\end{array}$$
+%%\vspace{-10pt}
+%\caption{\small Syntax of Core @$\lambda$.
+Here, $x$ ranges over external and internal language variables, $\tvar{t}$ ranges over type-level variables, $\fvar{tycon}$ ranges over type constructor names, $\ell$ ranges over labels, $n$ ranges over integers and $\pm$ ranges over standard binary operations over integers.
+The introductory form for arguments, $(\Gamma; e)?$, should only be constructed by the compiler, not by type constructor providers.
+%\label{grammar}}
+%\end{figure}
+\begin{figure*}[t]
+\small
+\begin{tabular}{r l l l}
+ & \textbf{initial} & \textbf{statically normal}\\
+\textbf{programs} & $\rho ::=$ & $\hat\rho ::=$\\
+ & ~~$\FF{using}~{\phi}~\FF{in}~e$ & ~~$\FF{using}~{\hat\phi}~\FF{in}~\hat e$\\
+\textbf{fragment decls.} & $\phi ::=$ & $\hat\phi ::=$\\
+tycon decls. & ~~$\FF{tycon}~\fvar{tycon} : \Theta = \theta$ & ~~$\FF{tycon}~\fvar{tycon} : \Theta = \hat\theta$\\
+static binding & ~~$\FF{def}~\tvar{x} : \kappa = \sigma$ & \\
+ & ~~$\phi; \phi$ & ~~$\hat\phi; \hat\phi$\\
+\textbf{static terms} & $\sigma ::=$ & $\hat\sigma ::= $ &\textbf{kinds}~~~  $\kappa ::= $\\
+static values & ~~$\hat \sigma$\\
+static variables & ~~$\tvar{x}$ & ~ & ~\\
+types & ~~$\ttype{tycon}{\sigma}$ & ~~$\ttype{tycon}{\hat\sigma}$ & ~~$\kTypeBlur$\\
+static products & ~~$\tunit$ & ~~$\tunit$ & ~~$\kunit$\\
+& ~~$(\sigma, \sigma)$ & ~~$(\hat\sigma, \hat\sigma)$ & ~~$\kpair{\kappa}{\kappa}$\\
+static sums & ~~$\FF{in}[\ell](\sigma)$ & ~~$\FF{in}[\ell](\hat\sigma)$ & ~~$\Sigma[\{\overline{\ell \hookrightarrow \kappa}\}]$\\
+static lists & ~~$[]_\kappa$ & ~~$[]_\kappa$ & ~~$\klist{\kappa}$\\
+ & ~~$\sigma :: \sigma$ & ~~$\hat\sigma :: \hat\sigma$ & \\
+static labels & ~~$\ell$ & ~~$\ell$ & ~~$\FF{L}$\\
+\\
+\textbf{external terms} & $e ::= $ & $\hat e ::=$ \\
+variables & ~~$x$ & ~~$x$\\
+binding & ~~$\FF{let}~x=e~\FF{in}~e$ & ~~$\FF{let}~x=\hat{e}~\FF{in}~\hat{e}$\\
+fixpoints & ~~$\FF{fix}~x{:}\sigma~\FF{is}~e$ & ~~$\FF{fix}~x{:}\hat\sigma~\FF{is}~\hat e$\\
+ascription & ~~$e : \sigma$ & ~~$\hat{e} : \hat\sigma$\\
+%introductory forms & ~~$\nu$ & ~~$\hat\nu$\\
+%eliminatory forms & ~~$\eta$ & ~~$\hat\eta$\\
+% & $\nu ::= $ & $\hat \nu ::=$\\
+analytic lambda & ~~$\lambda x.e$ & ~~$\lambda x.\hat{e}$\\
+%synthetic lambda & ~~$\lambda x {:} \sigma.e$ & ~~$\lambda x{:}\hat\sigma.\hat e$\\
+analytic intro & ~~$\FF{intro}[\sigma](\overline{e})$ & ~~$\FF{intro}[\hat\sigma](\overline{\hat e})$\\
+%synthetic intro & ~~$\fvar{tycon}\cdot\FF{intro}[\sigma](\overline{e})$ & ~~$\fvar{tycon}\cdot\FF{intro}[\hat \sigma](\overline{\hat e})$\\
+% & $\eta ::= $ & $\hat\eta ::=$\\
+synthetic elim & ~~$e\cdot\FF{elim}[\sigma](\overline{e})$ & ~~$\hat e\cdot\FF{elim}[\hat \sigma](\overline{\hat e})$\\
+\end{tabular}
+\caption{Syntax from the perspective of fragment clients (i.e. normal developers)}
+\end{figure*}
+\begin{figure*}[t]
+\small
+\hspace{-5px}
+$
+\begin{array}{rclr}
+%\FF{intro}(e_1; \ldots; e_n) & := & \FF{intro}[()](e_1; \ldots; e_n)\\
+\sigma\langle e_1; \ldots; e_n \rangle & := & \FF{intro}[\tfst{\sigma}](e_1; \ldots; e_n) : \tsnd{\sigma} & \text{assisted intro}\\
+%\langle\sigmaidx\rangle(e_1; \ldots; e_n) & := & \FF{intro}[\sigmaidx](e_1; \ldots; e_n)\\
+%[\sigmaidx](e_1; \ldots; e_n) & := & \FF{intro}[\sigmaidx](e_1; \ldots; e_n)\\
+\{ \ell_1{=}~e_1, \ldots, \ell_n{=}~e_n \} & := & \FF{intro}[\ell_1 :: \ldots :: \ell_n :: []_{\FF{L}}](e_1; \ldots; e_n) & \text{labeled seq.}  \\
+(e_1, ..., e_n) & := & \FF{intro}[()](e_1; \ldots; e_n) & \text{unlabeled seq.}\\
+%n & := & \FF{intro}[n]() & \text{number literal} \\
+%s & := & \FF{intro}[s]() & \text{string literal}\\
+%e\cdot\FF{elim}(e_1; \ldots; e_n) & := & e\cdot\FF{elim}[()](e_1; \ldots; e_n)\\
+e~e_1 & := & e\cdot\FF{elim}[()](e_1) & \text{application}\\
+\sigma~e~\langle e_1; \ldots; e_n \rangle  & := & e\cdot\FF{elim}[\sigma](e_1; \ldots; e_n) & \text{assisted elim}\\
+e.\ell & := & e\cdot\FF{elim}[\ell]() & \text{projection}\\
+\F{case}e~\{\ell_1\langle x_1 \rangle \Rightarrow e_1~|~\ldots~|~\ell_n\langle x_n \rangle \Rightarrow e_n\} & := & e\cdot\FF{elim}[\ell_1 :: \ldots :: \ell_n :: []_{\FF{L}}]( & \text{case analysis}\\
+ & & \quad \lambda x_1.e_1; \ldots; \lambda x_n.e_n)\\
+\sigma_1 \rightharpoonup \sigma_2 & := & \ttype{parr}{(\sigma_1, \sigma_2)} & \text{arrow types}
+%!TEX encoding = UTF-8 Unicodee.\ell(e_1; ...; e_n) & := & e\cdot\FF{elim}[\ell](e_1; \ldots; e_n)
+%(e_1, ..., e_n) & := & intro[()](e_1; ...; e_n)
+%n &  := & intro[n]()
+%s & := & intro[s]()
+%
+%e.l & := & elim[l](e)
+%e[e_1; ...; e_n] & := & elim[()](e_1, ..., e_n)
+%e.l(e_1; ...; e_n) & := & elim[l]
+\end{array}
+$
+\caption{\small Desugaring from conventional concrete syntax to core forms. The number and string literal forms assume type-level numbers, $n$, and strings, $s$ (details not shown).}
+\label{desugaring}
+\end{figure*}
+\newcommand{\cev}[1]{\reflectbox{\ensuremath{\vec{\reflectbox{\ensuremath{#1}}}}}}
+\newcommand{\tri}{\cev\iota}
+\newcommand{\trt}{\cev\tau}
+\begin{figure*}
+\newcommand{\tcl}[1]{\multicolumn{2}{l}{#1}}
+\small
+\begin{tabular}{r l l l l}
+ & \tcl{\textbf{initial}} & \textbf{statically normal} \\
+\textbf{tycon definitions} & \tcl{$\delta ::=$}  & $\hat\delta ::=$ \\
+ & \tcl{~~$\FF{iana}~\sigma; \FF{esyn}~\sigma; \FF{rep}~\sigma$} & ~~$\FF{iana}~\hat\sigma; \FF{esyn}~\hat\sigma; \FF{rep}~\hat\sigma$\\
+ \\
+\textbf{(static language, cont.)} & \tcl{$\sigma ::= ...$} & $\hat\sigma ::= ...$ & $\kappa ::= ...$\\
+type elim & \tcl{~~$\tfamcase{\sigma}{tycon}{x}{\sigma}{\sigma}$} & \\
+static product elim & \tcl{~~$\sigma.\FF{prl}$} &\\
+ & \tcl{~~$\sigma.\FF{prr}$} &\\
+static sum elim & \tcl{~~$\FF{case}~\sigma~\{\overline{l\langle\tvar{x}\rangle\Rightarrow\sigma}\}$} &\\
+static list elim & \tcl{~~$\FF{listrec}(\sigma; \sigma; \tvar{h}, \tvar{t}, \tvar{r}.\sigma)$} & ~\\
+decidable equality & \tcl{~~$\tifeq{\hat\sigma}{\hat\sigma}{\kappa}{\hat\sigma}{\hat\sigma}$} & ~ & ~\\
+type error & \tcl{~~$\FF{tyerr}$} & ~ \\
+static functions & \tcl{~~$\tlam{x}{\kappa}{\sigma}$} & ~~$\tlam{x}{\kappa}{\sigma}$ & ~~$\karrow{\kappa}{\kappa}$\\
+ & \tcl{~~$\sigma~\sigma$} & ~ & ~\\
+term translations & \tcl{~~$\titerm{\tri}$} & ~~$\titerm{\iota}$ & ~~$\kITerm$\\
+type translations & \tcl{~~$\titype{\trt}$} & ~~$\titype{\tau}$ & ~~$\kIType$\\
+ & \tcl{~~$\trepof{\sigma}$} & ~\\
+arguments & \multicolumn{3}{c}{\color{gray}(introduced only by typechecker)} & ~~$\Q$\\
+ & \tcl{~~$\FF{ana}(\sigma; \sigma; \tvar{x}.\sigma)$} & \\
+ & \tcl{~~$\FF{syn}(\sigma; \tvar{x}, \tvar{t}.\sigma)$} & \\
+\\
+\textbf{internal language} & $\tri ::= $ & $\trt ::= $ & $\iota ::= $ & $\tau ::=$\\
+splice & ~~$\iup{\sigma}$ & ~~$\dup{\sigma}$ & \\
+variables & ~~$x$ &  & ~~$x$ & \\
+fixpoints & ~~$\FF{fix}~x{:}\trt~\FF{is}~\tri$ & & ~~$\FF{fix}~x{:}\tau~\FF{is}~\iota$\\
+functions & ~~$\lambda x{:}\trt.\tri$ & ~~$\trt \rightharpoonup \trt$ & ~~$\lambda x{:}\tau.\iota$ & ~~$\tau \rightharpoonup \tau$\\
+ & ~~$\tri~\tri$ & ~ & ~~$\iota~\iota$\\
+products & ~~$()$ & ~~$\dunit$ & ~~$()$ & ~~$\dunit$\\
+ & ~~$(\tri, \tri)$ & ~~$\dpair{\trt}{\trt}$ & ~~$(\iota, \iota)$ & ~~$\dpair{\tau}{\tau}$\\
+ & ~~$\tri.\FF{prl}$ & ~ & ~~$\tri.\FF{prl}$ & ~\\
+  & ~~$\tri.\FF{prr}$ & ~ & ~~$\tri.\FF{prr}$ & ~\\
+%products & ~~$(\overline{l=\tri})$ & ~~$\Pi[\overline{(l, \trt)}]$ & ~~$(\overline{l = \iota})$ & ~~$\Pi[\overline{(l, \tau)}]$\\
+% & ~~$\tri\cdot l$ & ~ & ~~$\iota\cdot l$ & ~\\
+sums & ~~$\FF{in}[\ell](\tri)$ & ~~$\Sigma[\{\overline{\ell \hookrightarrow \trt}\}]$ & ~~$\FF{in}[\ell](\iota)$ & ~~$\Sigma[\{\overline{\ell \hookrightarrow \tau}\}]$\\
+ & ~~$\FF{case}~ \tri ~\{\overline{\ell(x) \Rightarrow \tri}\}$ & ~ & ~~$\FF{case}~\iota~\{\overline{\ell(x)\Rightarrow \iota}\}$ & ~\\
+integers & ~~$\FF{int}[z]$ & ~~$\FF{int}$ & ~~$\FF{int}[z]$ & ~~$\FF{int}$\\
+ & ~~$\tri \pm \tri$ & ~ & ~~$\iota \pm \iota$ & ~\\
+ & ~~$\FF{if0}(\tri; \tri; \tri; \tri)$ & ~ & ~~$\FF{if0}(\iota; \iota; \iota; \iota)$ & ~
+ %& ~~$
+\end{tabular}
+\caption{Syntax from the perspective of fragment providers also includes an internal language and static language.}
+\end{figure*}
+\begin{figure*}
+\small
+\begin{tabular}{r l l}
+external typing context & $\Gamma ::= \emptyset \pipe \Gamma, x \Rightarrow \ttype{tycon}{\hat\sigma}$\\
+\\
+ & \textbf{initial} & \textbf{statically normal} \\
+\textbf{static language, cont.} & $\sigma ::= ...$ & $\hat\sigma ::= ...$\\
+argument intro & ~~$\FF{arg}(\Gamma; \hat e)$ & ~~$\FF{arg}(\Gamma; \hat e)$
+\end{tabular}
+\caption{Syntax from the perspective of the semantics / compiler.}
+\end{figure*}
+\newcommand{\fkctx}{\mathcal{F}}
+\newcommand{\ok}[1]{#1~\mathtt{ok}}
+\newcommand{\normalize}[3]{#1 \curlyveedownarrow_{#2} #3}
+\newcommand{\normalizeX}[2]{\normalize{#1}{\Phi}{#2}}
+\newcommand{\serr}[1]{#1~\mathtt{tyerr}}
+\begin{figure*}
+\small
+\begin{tabular}{r l l}
+\textbf{context} & \textbf{syntax} & \textbf{well-formedness}\\
+fragment kind context & $\fkctx ::= \fkctx_0 \pipe \fkctx, \fvar{tycon}[\kappa]~\{\FF{intro}[\kappa]; \FF{elim}[\kappa]\}$ & $\vdash \fkctx$\\
+kinding context & $\Delta ::= \Delta \pipe \Delta, \tvar{x} : \kappa$\\
+fragment context & $\Phi ::= \Phi_0 \pipe \Phi, \fvar{tycon}~\{\hat\delta\}$ & $\vdash \Phi \sim \fkctx$\\
+external typing context & $\Gamma ::= \emptyset \pipe \Gamma, x \Rightarrow \hat\sigma$ & $\vdash_\Phi \Gamma$\\
+internal typing context & $\Omega ::= \emptyset \pipe \Omega, x : \tau$\\
+~
+\end{tabular}
+\\
+\\
+\begin{tabular}{l l l l l}
+\textbf{kinding}& \multicolumn{2}{l}{\textbf{static normalization}} & \multicolumn{2}{l}{\textbf{program compilation}}\\
+$\ok{\rho}$ & $\normalize{\rho}{}{\hat\rho}$ & $\serr{\rho}$ & $\hat\rho \leadsto \iota$\\
+$\Delta \vdash_\fkctx {\phi}\sim\fkctx$ & $\normalizeX{\phi}{\hat\phi}$ & $\serr{\phi}$ & $\hat\phi \sim \Phi$\\
+$\Delta \vdash_\fkctx \delta\sim\{\kappa; \kappa\}$ & $\normalizeX{\delta}{\hat\delta}$ & $\serr{\delta}$ & \multicolumn{2}{l}{\textbf{active typechecking and translation}}\\
+$\Delta \vdash_\fkctx \sigma : \kappa$ & $\normalizeX{\sigma}{\hat \sigma}$ & $\serr{\sigma}$ & $\Gamma \vdash_\Phi \hat e \Leftarrow \hat\sigma \leadsto \iota$ & $\Gamma \vdash_\Phi \hat e \Rightarrow \hat\sigma \leadsto \iota$\\
+$\Delta \vdash_\fkctx \ok{e}$ & $\normalizeX{e}{\hat e}$ & $\serr{e}$ & $\vdash_\Phi \hat\sigma \leadsto \tau$ & $\vdash_\Phi \Gamma \leadsto \Omega$\\
+$\Delta \vdash_\fkctx \ok{\tri}$ & $\normalizeX{\tri}{\iota}$ & $\serr{\tri}$ & \multicolumn{1}{l}{\textbf{internal statics}} & \textbf{internal dynamics}\\
+$\Delta \vdash_\fkctx \ok{\trt}$ & $\normalizeX{\trt}{\tau}$ & $\serr{\trt}$ & $\Omega \vdash \iota : \tau$ & $\iota \mapsto \iota$~~~~~~~~$\iota~\mathtt{val}$
+\end{tabular}
+\caption{Summary of judgements}
+\end{figure*}
+\begin{figure*}
+\small
+$
+\begin{array}{lcl}
+\delta_{\texttt{nat}} & := & \quad \FF{iana}~\tlam{opidx}{\kunit+\kunit}{\tlam{tyidx}{\kunit}{\tlam{args}{\klist{\FF{Arg}}}{
+	\tsumcase{\tvar{opidx}\\
+&&\quad\quad}{\_}{\tvar{arity0}~\tvar{args}~\titerm{0}\\
+&&\quad\quad}{\_}{\tvar{arity1}~\tvar{args}~\tlam{a}{\FF{Arg}}{\FF{ana}(\tvar{a}; \ttype{nat}{\tunit}; \tvar{x}.\titerm{\iup{\tvar{x}}+1})}}
+}}}\\
+&&\quad \FF{esyn}~\tlam{opidx}{\kunit}{\tlam{tyidx}{\kunit}{\tlam{x}{\kITerm}{\tlam{args}{\klist{\FF{Arg}}}{\tvar{arity2}~\tvar{args}~\tlam{a1}{\FF{Arg}}{\tlam{a2}{\FF{Arg}}{\\
+&&\quad\quad \FF{syn}(\tvar{a1}; \tvar{t1}, \tvar{x1}.\\
+&&\quad\quad\quad \F{let}\tvar{t2} : \kTypeBlur = \ttype{arrow}{(\ttype{nat}{\tunit}, \ttype{arrow}{(\tvar{t1}, \tvar{t1})})}~\F{in} \\
+&&\quad\quad\quad \FF{ana}(\tvar{a2}; \tvar{t2}; \tvar{x2}.(\tvar{t1},\\
+&&\quad\quad\quad\quad \titerm{\iapp{(\ifix{f}{\darrow{\dint}{\dup{\trepof{\tvar{t2}}}}}{\ilam{x}{\dint}{\\&&\quad\quad\quad\quad\quad\quad \FF{if0}(x; \iup{\tvar{x1}}; {
+		%\\&\quad\quad\quad\quad\quad\quad
+		\iapp{\iup{\tvar{x2}}~(x-1)}{(\iapp{f}{(x-1)})}})}}\\
+		&&\quad\quad\quad\quad\quad)}{\iup{\tvar{x}}}})))}}}}}}\\
+&&\quad \FF{rep}~\tlam{tyidx}{\kunit}{\titype{\dint}}\\
+\end{array}
+$
+\caption{\small The definition of the nat type constronstructor.}
+\end{figure*}
+%\begin{figure}
+%\small
+%\begin{flalign}
+%& \F{tycon}\fvar{record}~\F{of}\klist{\kpair{\FF{L}}{\kTypeBlur}}~\{\\
+%& \quad \FF{iana}~\{\tlam{i}{\kpair{\klist{\kpair{\FF{L}}{\kTypeBlur}}}{\klist{\FF{L}}}}{\tlam{a}{\klist{\Q}}{\\& \quad\quad\tfold{\tvar{zip3}~\tfst{\tvar{i}}~\tsnd{\tvar{i}}~\tvar{a}}
+%	 {\titerm{\tunit}}{h}{t}{r}{
+%	\\&\quad\quad\quad\tifeq{\tfst{\tvar{first}~\tvar{h}}}{\tvar{second}~\tvar{h}}{\FF{L}}{
+%	\\&\quad\quad\quad\quad
+%	\F{let}~\tvar{x}:\kITerm~=~\FF{ana}(\tvar{third}~\tvar{h}; \tsnd{\tvar{first}~\tvar{h}})
+%~\FF{in}
+%\\ & \quad\quad\quad\quad \tfold{\tvar{t}}{\tvar{x}
+%}{\_}{\_}{\_}{\titerm{(\iup{\tvar{x}},\iup{\tvar{r}})}}	}{\terr}
+%%     \F{let}~\tvar{tt}:\FF{TT}=~\FF{syn}(\tsnd{\tsnd{\tvar{h}}}) \F{in}\\
+%%\\     & \quad\quad\quad 
+%}
+%}}\}\\
+%& \quad \FF{isyn}~\{\tlam{i}{\klist{\FF{L}}}{\tlam{a}{\klist{\Q}}{
+%\\&\quad\quad \tfold{\tvar{zip2}~\tvar{i}~\tvar{a}}{\tden{\titerm{()}}{\ttype{record}{\tnil{{\kpair{\FF{L}}{\kTypeBlur}}}}}}{h}{t}{r}{
+%\\&\quad\quad\quad \F{let}~\tvar{htt}:\FF{TT}~=~\FF{syn}(\tsnd{\tvar{h}})
+%~\FF{in}
+%\\&\quad\quad\quad \F{let}~\tvar{hty}:\kTypeBlur~=~\ttypeof{\tvar{htt}}
+%\\&\quad\quad\quad \F{let}~\tvar{ty}:\kTypeBlur~=~\ttype{record}{(\tfst{\tvar{h}}, \tvar{hty})::\tnil{{\kpair{\FF{L}}{\kTypeBlur}}}}
+%~\FF{in}
+%\\&\quad\quad\quad \tfold{\tvar t}{\tden{\itransof{\tvar{htt}}}{\tvar{ty}}}{\_}{\_}{\_}{
+%\\&\quad\quad\quad\quad \tden{\titerm{(\iup{\itransof{\tvar{htt}}},\iup{\itransof{\tvar{r}}})}}{\tvar{ty}}}}
+%}}\}\\
+%& \quad \FF{esyn}~\{\tlam{i}{\FF{L}}{\tlam{a}{\klist{\Q}}{
+%	\tvar{arity1}~\tvar{a}~\tlam{ty}{\kTypeBlur}{\tlam{x}{\kITerm}{
+%\\&\quad\quad		\tfamcase{\tvar{ty}}{record}{sig}{
+%\\&\quad\quad\quad \tfold{\tvar{sig}}{\terr}{h}{t}{r}{
+%\\&\quad\quad\quad\quad \tifeq{\tfst{\tvar{h}}}{\tvar{i}}{\FF{L}}{\tfold{\tvar{t}}{\tden{\tvar{x}}{\tsnd{\tvar{h}}}}{\_}{t'}{\_}{
+%\\&\quad\quad\quad\quad\quad \tfold{\tvar{t'}}{\tden{\titerm{\ifst{\iup{\tvar{x}}}}}{\tsnd{\tvar{h}}}}{\_}{\_}{\tvar{r'}}{
+%\\&\quad\quad\quad\quad\quad\quad \tden{\titerm{\isnd{\iup{{\itransof{\tvar r}}}}}}{\tsnd{\tvar{h}}}}}\\&\quad\quad\quad\quad\hspace{-3px}}{\tvar{r}}
+%}\\&\quad\quad\hspace{-3px}}{\terr}
+%	}}
+%}}\\
+%& \quad \FF{rep}~\{\tlam{i}{\klist{\kpair{\FF{L}}{\kTypeBlur}}}{ 	\\& \quad \quad \tfold{\tvar{i}}{\titype{\dunit}}{s}{j}{r}{
+% 		\tfold{\tvar{j}}{\trepof{\tsnd{\tvar{s}}}}{\_}{\_}{\_}{\\
+%		&\quad\quad\quad
+% 		\titype{\dpair{\dup{\trepof{\tsnd{\tvar{s}}}}}{\dup{\tvar{r}}}}
+% 		}
+% 	}}\}\\
+%& \}; \\
+%& \F{let}\tvar{R} : \kTypeBlur = \ttype{record}{(\ell_1, \ttype{record}{\tnil{{\FF{L}\times\kTypeBlur}}}) :: \tnil{{\FF{L}\times\kTypeBlur}}}~\F{in}\\
+%&\F{let}id = \elam{x}{\tvar{R}}{\{\ell_1=~x\cdot\ell_1\} :: \fvar{record}}\\
+%&\F{let}triv = id[\{\ell_1 = \{\}\}]\cdot\ell_1
+%%&\F{let}one = \FF{intro}[\ell_\text{succ}]
+%%&\F{let}plus = \elam{x}{\tvar{nat}}{\elam{y}{\tvar{nat}}{\FF{elim}[()](x; y; \\
+%%& \quad \elam{r}{\tvar{nat}}{intro[\ell_\text{succ}](r) : \tvar{nat}}})}
+%\end{flalign}
+%\caption{The definition of the record type in $\lamAce$ using the desugarings in Figure \ref{desugaring} and with the addition of simple let bindings for both type and term variables (not shown).
+Some type-level helper functions also omitted for concision.}
+%\label{record-theory}
+%\end{figure}
+
 %These correspond to the premises of the \emph{central compilation judgement} $\pkcompiles{\rho}{\iota}$:
@@
 %This anchors our exposition; we will describe how it is derived (i.e. how to write a compiler for $\lamAce$) in the following sections. 
-The key judgements are the \emph{bidirectional active typing judgements}  (Fig.
-\ref{att}, which we describe starting in Sec.
-\ref{opcons}).
-They relate an external term, $e$, to a {type}, $\tau$, called its \emph{type assignment}, and an internal term, $\iota$, called its \emph{translation}, under \emph{typing context} $\Gamma$ and \emph{constructor context} $\Phi$.
-The first is synthesis, the second analysis.
+The key judgements are the \emph{bidirectional active typechecking and translation judgements}, relating an external term, $e$, to a {type}, $\tau$, and an internal term, $\iota$, called its \emph{translation}, under \emph{typing context} $\Gamma$ and \emph{constructor context} $\Phi$.
 \[\Gamma \vdash_\fvalCtx e \Rightarrow \tau \leadsto \iota
-~~~~~~~
+~~~~~~~\text{and}~~~~~~~
 \Gamma \vdash_\fvalCtx e \Leftarrow \tau \leadsto \iota\]
@@
 The typing context, $\Gamma$, maps variables to types in essentially the conventional way (\cite{pfpl} contains the necessary background for this section).
-The constructor context, $\Phi$, tracks user-defined type  constructors.
+The constructor context, $\Phi$, tracks user-defined type  constructors introduced in the ``imported'' fragments.
+%Note that fragments can also export type-level values bound to type-level variables (written in bold font, e.g.
+$\tvar{nat}$, $\tvar{s}$, $\tvar{z}$, etc.), but substitution will be performed prior to typechecking, so they need not be tracked by $\Phi$.
 % Weakening and exchange of the constructor context closely related to the issue of conservativity that we will return to.
 Each constructor is identified by name, so there is no analog to contraction.
-
-The dynamic behavior of an external term is determined entirely by its translation to the internal language, which has a conventional operational semantics.
-This form of semantics can be seen as lifting into the language specification the first stage of a type-directed compiler like the TIL compiler for Standard ML \cite{tarditi+:til-OLD} and has some parallels to the Harper-Stone semantics for Standard ML, where external terms were also given meaning by elaboration from the EL to an IL \cite{Harper00atype-theoretic}.
-%
-
-In $\lamAce$, the internal language (IL) provides partial functions (via the generic fixpoint operator of Plotkin's PCF) and simple product  types for the sake of our example.
+ This form of semantics can be seen as lifting into the language specification the first stage of a type-directed compiler like the TIL compiler for Standard ML \cite{tarditi+:til-OLD} and has some parallels to the Harper-Stone semantics for Standard ML \cite{Harper00atype-theoretic}.
+There, as in Wyvern, external terms were given meaning by elaboration from the EL to an IL having the same type system.
+Here the two languages have different type systems, so we call it a translation rather than an elaboration (arranging the judgement form slightly differently to emphasize this distinction, cf.
+above).
+
+In @$\lambda$, the internal language (IL) provides partial functions (via the generic fixpoint operator of Plotkin's PCF), simple product  types and integers for the sake of our example (and as a nod toward practicality on contemporary machines).
 In practice, the internal language could be any typed  language with a specification for which type safety and decidability of typechecking have been satisfyingly determined.
+In Sec.
+\ref{ace}, we will see how the internal language can itself be made user-definable.
 The internal type system serves as a ``floor'': guarantees that must hold for terms of any type (e.g.
@@
 
+The external language has a fixed syntax with six forms: variables, $\FF{let}$-bindings, lambda terms, type ascription and generalized introductory and elimination forms.
+As we will see, the generalized introductory form is given meaning by the type it is analyzed against (similar to the protocol for TSLs in Wyvern), and the elimination form is given meaning by the type of the external term being eliminated ($e$).
+This represents an internalization into the language of Gentzen's inversion principle \cite{gentzen}\todo{what to cite for this?}.
+Fig.
+\ref{desugaring} shows how to recover more conventional introductory and elimination forms by a purely syntactic desugaring.%\footnote{Note that for even more flexibility, we could also include TSLs, but we choose to avoid that for simplicity (and to show that, strictly speaking, one need not have an extensible syntax to have an extensible type system, and \textit{vice versa}).}
+
 \subsection{Types and Type-Level Computation}\label{types}
-$\lamAce$ supports, and makes extensive use of, simply-kinded type-level computation.
+@$\lambda$ supports, and makes extensive use of, simply-kinded type-level computation.
 Specifically, type-level terms, $\tau$, themselves form a typed lambda calculus.
-The classifiers of type-level terms are called \emph{kinds}, $\kappa$, to distinguish them from  \emph{types}.
-Types are  type-level values of kind $\kTypeBlur$.
-In Ace, the type-level language (together with the level of programs) is written in Python, which can be thought of as having a rather limited kind system (with one kind, \verb|dyn|).
+The classifiers of type-level terms are called \emph{kinds}, $\kappa$, to distinguish them from  \emph{types}, which are  type-level values of kind $\kTypeBlur$.
+%In Ace, the type-level language (together with the level of programs) is written in Python, which can be thought of as having a rather limited kind system (with one kind, \verb|dyn|).
 Here, we are able to more precisely discuss the kinds of values in the type-level language.
-
-In most languages, types are formed by applying one of a collection of \emph{type constructors} to zero or more \emph{indices}.
-In $\lamAce$, the situation is notionally similar.
-User-defined type constructors can be declared at the top of a program (or lifted to the top, in practice) using \textsf{tycon}.
+As in Sec.
+\ref{att}, types are formed by applying a \emph{type constructor} to an \emph{index}.
+User-defined type constructors are declared in fragment definitions using $\FF{tycon}$.
 Each constructor in the program must have a unique name, written e.g.
-\fvar{record}.\footnote{We assume naming conflicts can be avoided by some extrinsic mechanism.} A type constructor must also declare an \emph{index kind}, $\kappaidx$.
+\fvar{nat} or \fvar{lprod}.
+%\footnote{We assume naming conflicts can be avoided by some extrinsic mechanism.}
+A type constructor must also declare an \emph{index kind}, $\kappat{tyidx}$.
+A type is introduced by applying a type constructor to an index of this kind, written $\ttype{Tycon}{\taut{tyidx}}$.
+For example, the type of natural numbers is indexed trivially (i.e.
+by kind $\kunit$), so it is written $\ttype{Nat}{\tunit}$.
 
  To permit the embedding of interesting type systems, the type-level language includes several kinds other than $\kTypeBlur$.
-We functional data structures to the type level: here, only unit ($\kunit$), binary products ($\kpair{\kappa_1}{\kappa_2}$) and lists ($\klist{\kappa}$), in addition to labels (introduced as $\ell$, possibly with a subscript, having kind \FF{L}).
-Our record type constructor is indexed by a list of pairs of {labels}  and types (a signature, in essence; line 1).
-The type constructor $\fvar{arrow}$ is included in the initial constructor context, $\fvalCtx_0$ and has index kind $\kpair{\kTypeBlur}{\kTypeBlur}$.
- In practice, one could include a richer functional programming language and retain the spirit of the calculus, as long as it does not introduce general recursion at the type level.
-
-A type is introduced by applying a type constructor to an index of this kind, written $\ttype{Tycon}{\tauidx}$.
-For example, the type of natural numbers is indexed trivially, so it is written $\ttype{Nat}{\tunit}$.
-We see a record type, abbreviated $\tvar{R}$, constructed on line 26.
-
-The kind $\kTypeBlur$ also has an elimination form, $$\tfamcase{\tau}{Tycon}{x}{\tau_1}{\tau_2}$$ allowing the extraction of a type index by case analysis against a contextually-available type constructor.
-To a first approximation, one might think of type constructors as constructors of a built-in open \cite{conf/ppdp/LohH06}, $\kTypeBlur$, at the type-level.
+We lift several functional data structures to the type level: here, only unit ($\kunit$), binary products ($\kpair{\kappa_1}{\kappa_2}$), binary sums ($\ksum{\kappa_1}{\kappa_2}$) and lists ($\klist{\kappa}$), in addition to labels (introduced as $\ell$, possibly with a subscript, having kind \FF{L}).
+The type constructor $\fvar{nat}$ is indexed trivially because there is only one natural number type, but $\fvar{lprod}$ would be indexed by a list of pairs of {labels}  and types.
+The type constructor $\fvar{arrow}$ is included in the initial constructor context, $\fvalCtx_0$, and has index kind $\kpair{\kTypeBlur}{\kTypeBlur}$.
+ As with the internal language, in practice, one could include a richer programming language and retain the spirit of the calculus, as long as it does not introduce general recursion at the type level.
+For example, our desugarings add support for number literals and string literals, which require adding numbers and strings to the type-level language (and providing a means for lifting them from the type-level language to the internal language, as we will discuss).
+
+ %We see a record type, abbreviated $\tvar{R}$, constructed on line 26.
+
+The kind $\kTypeBlur$ also has an elimination form, $\tfamcase{\tau}{Tycon}{x}{\tau_1}{\tau_2}$ allowing the extraction of a type index by case analysis against a contextually-available type constructor.
+To a first approximation, one might think of type constructors as constructors of a built-in open datatype \cite{conf/ppdp/LohH06}, $\kTypeBlur$, at the type-level.
 Like open datatypes, there is no notion of exhaustiveness so the default case is required for totality.
-We will see where this is used shortly.
+%We will see where this is used shortly.
 
@@
 The type-level language does, however, include total functions of arrow kind, written $\karrow{\kappa_1}{\kappa_2}$.
-Type constructor application can be wrapped in a type-level function to emulate a first-class or uncurried version of a type constructor for convenience (indeed, such a wrapper could be generated automatically, though we do not do so).
+Type constructor application can be wrapped in a type-level function to emulate a first-class or uncurried version of a type constructor for convenience.% (indeed, such a wrapper could be generated automatically, though we do not do so).
 
@@
 
-Every type constructor also defines a \emph{representation}, a type-level function that associates with every type an internal \emph{representation type} (analagous to \verb|trans_type| above).
-We will return to this after introducing the external forms and operator definitions.
-
-\subsection{Core External Forms and Desugaring}\label{opcons}
-The syntax for external terms (Figure \ref{grammar}) contains variables, $\lambda$ terms, three generalized introductory forms and a single generalized elimination form.
+%Every type constructor also defines type-level functions called $\FF{iana}$ and $\FF{esyn}$, which we will describe below, and a \emph{representation schema}, a type-level function that associates with every type an internal type.
+We will return to this after introducing operators.
+%
+%\subsection{Core External Forms and Desugaring}\label{opcons}
+%The syntax for external terms (Figure \ref{grammar}) contains variables, $\lambda$ terms, three generalized introductory forms and a single generalized elimination form.
 The introductory forms are either unascribed, ascribed with a type or ascribed with a type constructor, as in our discussion of Ace.
@@
 It is instructive to rewrite lines 27-28 of Figure \ref{record-theory} using these desugarings.% In Ace, desugarings can be user-defined
-\begin{figure}[t]
-\small
-\[
-\begin{array}{rcl}
-\{ \ell_1{=}~e_1, \ldots, \ell_n{=}~e_n \} & := & \FF{intro}[\ell_1 :: \ldots :: \ell_n :: []](e_1; \ldots; e_n)\\
-(e_1, ..., e_n) & := & \FF{intro}[()](e_1; \ldots; e_n)\\
-\#n & := & \FF{intro}[n]()\\
-"s" & := & \FF{intro}[s]()\\
-e.\ell & := & e\cdot\FF{elim}[\ell]()\\
-e[e_1; \ldots; e_n] & := & e\cdot\FF{elim}[()](e_1; \ldots; e_n)\\
-e.\ell(e_1; ...; e_n) & := & e\cdot\FF{elim}[\ell](e_1; \ldots; e_n)
-%(e_1, ..., e_n) & := & intro[()](e_1; ...; e_n)
-%n &  := & intro[n]()
-%s & := & intro[s]()
-%
-%e.l & := & elim[l](e)
-%e[e_1; ...; e_n] & := & elim[()](e_1, ..., e_n)
-%e.l(e_1; ...; e_n) & := & elim[l]
-\end{array}
-\]
-\caption{Desugaring from conventional syntax to core forms. We assume that the type-level language has numbers, $n$, and strings, $s$ (not shown for concision).}
-\label{desugaring}
-\end{figure}
-
+%
 %User-defined operator constructors are declared using \textsf{opcon}.
@@
 
-\subsection{Operator Definitions and Representational Consistency}
+\subsection{Bidirectional Active Typechecking and Translation}
 
@@
 \newcommand{\atjanaX}[3]{\Gamma \vdash_\fvalCtx #1 \Leftarrow #2 \leadsto #3}
+\newcommand{\atjerrX}[1]{\Gamma \vdash_\fvalCtx #1~ \mathtt{error}}
 \begin{figure}[t]
 \small
-$\fbox{\inferrule{}{\ecompilesAX{e}{\tau}{\iota}}}$
-~~~~$\Gamma ::= \emptyset \pipe \Gamma, x \Rightarrow \tau$
+$\fbox{\inferrule{}{\atjsynX{e}{\tau}{\iota}}}$~~~~
+$\fbox{\inferrule{}{\atjanaX{e}{\tau}{\iota}}}$~~~~
+%$\fbox{\inferrule{}{\atjerrX{e}}}$
 \begin{mathpar}
+\inferrule[att-flip]{
+	\atjsynX{e}{\tau}{\iota}
+}{
+	\atjanaX{e}{\tau}{\iota}
+}
+
 \inferrule[att-var]{
@@
 
-\inferrule[att-syn-to-ana]{
-	\atjsynX{e}{\tau}{\iota}
+\inferrule[att-asc]{
+    \tau \Downarrow_\fvalCtx \tau'\\
+	\atjanaX{e}{\tau'}{\iota}
 }{
-	\atjanaX{e}{\tau}{\iota}
+	\atjsynX{e : \tau}{\tau'}{\iota}
 }
 
-\inferrule[att-lam]{
-	\tau_1 \Downarrow_\fvalCtx \tau_1'\\
-		\trepof{\tau_1'} \Downarrow_\fvalCtx \titype{\sigma}\\\\
-	\Gamma, x \Rightarrow \tau_1' \vdash_\fvalCtx e \Rightarrow \tau_2 \leadsto \iota
-%	\iota \hookrightarrow_\fvalCtx \iota'\\
-%		\sigma \hookrightarrow_\fvalCtx \sigma'
-%	\ddbar{\fvar{Arrow}}{\fvalCtx}{\trepof{\tau_1'}}{\sbar_1}\\
-	%\delfromtau{$\Xi_0$}{\fvalCtx}{\tau_1'}{\sabs}\\\\
+\inferrule[att-let-syn]{
+	\atjsynX{e_1}{\tau_1}{\iota_1}\\
+	\Gamma, x \Rightarrow \tau_1 \vdash_\fvalCtx e_2 \Rightarrow \tau_2 \leadsto \iota_2\\
+	\trepof{\tau_1} \Downarrow_\fvalCtx \titype{\sigma_1}
 }{
-	\atjsynX{\elam{x}{\tau_1}{e}}{\ttype{arrow}{(\tau_1', \tau_2)}}{\ilam{x}{\sigma'}{\iota}}
+	\atjsynX{\F{let}x = e_1~\F{in}e_2}{\tau_2}{(\ilam{x}{\sigma_1}{\iota_2})~\iota_1}
 }
 
-\inferrule[att-i-unasc]{
+\inferrule[att-lam-ana]{
+	\Gamma, x \Rightarrow \tau_1 \vdash_\fvalCtx e \Leftarrow \tau_2 \leadsto \iota\\
+	\trepof{\tau_1} \Downarrow_\fvalCtx \titype{\sigma_1}
+}{
+	\atjanaX{\lambda x.e}{\ttype{arrow}{(\tau_1, \tau_2)}}{\ilam{x}{\sigma_1}{\iota}}
+}
+
+%\inferrule[att-lam]{
+%	\tau_1 \Downarrow_\fvalCtx \tau_1'\\
+%		\trepof{\tau_1'} \Downarrow_\fvalCtx \titype{\sigma}\\\\
+%	\Gamma, x \Rightarrow \tau_1' \vdash_\fvalCtx e \Rightarrow \tau_2 \leadsto \iota
+%%	\iota \hookrightarrow_\fvalCtx \iota'\\
+%%		\sigma \hookrightarrow_\fvalCtx \sigma'
+%%	\ddbar{\fvar{Arrow}}{\fvalCtx}{\trepof{\tau_1'}}{\sbar_1}\\
+%	%\delfromtau{$\Xi_0$}{\fvalCtx}{\tau_1'}{\sabs}\\\\
+%}{
+%	\atjsynX{\elam{x}{\tau_1}{e}}{\ttype{arrow}{(\tau_1', \tau_2)}}{\ilam{x}{\sigma'}{\iota}}
+%}
+%
+\inferrule[att-intro-ana]{
 	\vdash_\fvalCtx \FF{iana}(\fvar{tycon})=\taudef\\
-	\taudef~(\tauidx, \tauidx')~((\Gamma; e_1)? :: \ldots :: (\Gamma; e_n)? :: []) \Downarrow_\fvalCtx \titerm{\iota}\\
+	\taudef~\taut{opidx}~\taut{tyidx}~((\Gamma; e_1)? :: \ldots :: (\Gamma; e_n)? :: []) \Downarrow_\fvalCtx \titerm{\iota}\\
 %	\trepof{\ttype{tycon}{\tauidx'}} \Downarrow_\fvalCtx \titype{\sigma}\\
-		\trepof{\ttype{tycon}{\tauidx'}} \Downarrow_\fvalCtx \titype{\sigma}\\
+		\trepof{\ttype{tycon}{\taut{tyidx}}} \Downarrow_\fvalCtx \titype{\sigma}\\
 		\Gamma \vdash_\fvalCtx \iota : \sigma
 }{
-	\atjanaX{\FF{intro}[\tauidx](e_1; \ldots; e_n)}{\ttype{tycon}{\tauidx'}}{\iota}
+	\atjanaX{\FF{intro}[\taut{opidx}](e_1; \ldots; e_n)}{\ttype{tycon}{\taut{tyidx}}}{\iota}
 }
 
-\inferrule[att-i-asc-ty]{
-	\atjanaX{I}{\tau}{\iota}
+%\inferrule[att-i-asc-ty]{
+%	\atjanaX{I}{\tau}{\iota}
+%}{
+%	\atjsynX{I : \tau}{\tau}{\iota}
+%}
+%
+%\inferrule[att-i-asc-tycon]{
+%	\vdash_\fvalCtx \FF{isyn}(\fvar{tycon})=\taudef\\\\
+%	\taudef~\tauidx~((\Gamma; e_1)? {::}{\ldots}{::}(\Gamma; e_n)? {::} []) \Downarrow_\fvalCtx \tden{\titerm{\iota}}{\ttype{tycon}{\tauidx'}}\\
+%			\trepof{\tau_1'} \Downarrow_\fvalCtx \titype{\sigma}\\
+%	\Gamma \vdash_\fvalCtx \iota : \sigma
+%}{
+%	\atjsynX{\FF{intro}[\tauidx](e_1; \ldots; e_n)] :: \fvar{tycon}}{\ttype{tycon}{\tauidx'}}{\iota'}
+%}
+%
+\inferrule[att-elim-syn]{
+	\atjsynX{e}{\ttype{tycon}{\taut{tyidx}}}{\iota}\\
+	\vdash_\fvalCtx \FF{esyn}(\fvar{tycon})=\taudef\\\\
+	\taudef~\taut{opidx}~\taut{tyidx}~\titerm{\iota}~((\Gamma; e)? :: (\Gamma; e_1)? :: \ldots :: (\Gamma; e_n)? :: []) \Downarrow_\fvalCtx (\tau, \titerm{\iota'})\\
+			\trepof{\tau} \Downarrow_\fvalCtx \titype{\sigma}\\
+	\Gamma \vdash_\fvalCtx \iota' : \sigma
 }{
-	\atjsynX{I : \tau}{\tau}{\iota}
-}
-
-\inferrule[att-i-asc-tycon]{
-	\vdash_\fvalCtx \FF{isyn}(\fvar{tycon})=\taudef\\\\
-	\taudef~\tauidx~(\Gamma; e_1)? {::}{\ldots}{::}(\Gamma; e_n)? {::} []) \Downarrow_\fvalCtx \tden{\titerm{\iota}}{\ttype{tycon}{\tauidx'}}\\
-			\trepof{\tau_1'} \Downarrow_\fvalCtx \titype{\sigma}\\
-	\Gamma \vdash_\fvalCtx \iota : \sigma
-}{
-	\atjsynX{\FF{intro}[\tauidx](e_1; \ldots; e_n)] :: \fvar{tycon}}{\tau}{\iota'}
-}
-
-\inferrule[att-elim]{
-	\atjsynX{e}{\ttype{tycon}{-}}{-}\\
-	\vdash_\fvalCtx \FF{esyn}(\fvar{tycon})=\taudef\\
-	\taudef~((\Gamma; e)? :: (\Gamma; e_1)? :: \ldots :: (\Gamma; e_n)? :: []) \Downarrow_\fvalCtx \tden{\tau}{\iota}\\
-			\trepof{\tau} \Downarrow_\fvalCtx \titype{\sigma}\\
-	\Gamma \vdash_\fvalCtx \iota : \sigma
-}{
-	\atjsynX{e\cdot\FF{elim}[\tauidx](e_1; \ldots; e_n)}{\tau}{\iota}
+	\atjsynX{e\cdot\FF{elim}[\taut{opidx}](e_1; \ldots; e_n)}{\tau}{\iota'}
 }
 \end{mathpar}
-\caption{\small The active typing judgement. The normalization judgement for type-level terms ($\Downarrow$) and the representational consistency check will be in an appendix.}
+\vspace{-10px}
+\caption{\small The bidirectional active typechecking and translation judgements.}
 \label{atj}
 \end{figure}
-\begin{figure}
+\begin{figure}[t]
 \small
+$\fbox{\inferrule{}{\tau \Downarrow_\fvalCtx \tau'}}$~~~~
 \begin{mathpar}
@@
 	\vdash_\fvalCtx \FF{rep}(\fvar{tycon}) = \taurep\\
-	\taurep \tauidx \Downarrow_\fvalCtx \titype{\sigma}
+	\taurep~\tauidx \Downarrow_\fvalCtx \titype{\sigma}
 }{
@@
     \tau \Downarrow_\fvalCtx (\Gamma; e)?\\
-	\atjsynX{e}{\tau}{\iota}
+	\atjsynX{e}{\tau}{\iota}\\
+    [\tau/\tvar{t}_{ty}, \titerm{\iota}/\tvar{t}_{trans}]\tau_2 \Downarrow_\fvalCtx \tau_2'
 }{
-	\FF{syn}(\tau) \Downarrow_\fvalCtx \tden{\titerm{\iota}}{\tau}
+	\FF{syn}(\tau_1; \tvar{t}_{ty}, \tvar{t}_{trans}.\tau_2) \Downarrow_\fvalCtx \tau_2'
 }
@@
 	\tau_1 \Downarrow_\fvalCtx (\Gamma; e)?\\
-	\tau_2 \Downarrow_\fvalCtx \ttype{tycon}{\tauidx}\\\\
-	\atjanaX{e}{\ttype{tycon}{\tauidx}}{\iota}\\
+	\tau_2 \Downarrow_\fvalCtx \tau_2'\\
+	\atjanaX{e}{\tau_2'}{\iota}\\
+	[\titerm{\iota}/\tvar{t}_{trans}]\tau_3 \Downarrow_\fvalCtx \tau_3'
 }{
-	\FF{ana}(\tau_1; \tau_2) \Downarrow_\fvalCtx \titerm{\iota}
+	\FF{ana}(\tau_1; \tau_2; \tvar{t}_{trans}.\tau_3) \Downarrow_\fvalCtx \tau_3'
 }
 \end{mathpar}
-\caption{\small Evaluation semantics for the type-level language.
-Missing rules (including error propagation rules, which immediately cause failure of typechecking) are unsurprising and will be given in an appendix.}
+\caption{\small Normalization semantics for the type-level language.
+Missing rules (including error propagation rules and normalization of quoted internal terms and types) are unsurprising and will be given later.}
 \label{tleval}
 \end{figure}
-The expressive and metatheoretic power of the calculus arises from how the rules for the active typing judgement handle these  generalized forms.
+The rules for bidirectional active typechecking and translation are shown in Fig. \ref{atj}.
+
+%The expressive and metatheoretic power of the calculus arises from how the rules for the active typing judgement handle these  generalized forms.
 Rather than fixing the specification of a finite collection of operator constructors and tasking the \emph{compiler} with deciding a typing derivation on its basis, the specification instead delegates to a type-level function associated with a type constructor.
@@
 This is a minor inconvenience in some cases that is resolved in Ace by the use of methods.} We see how this is done with the rules for the active typing judgements, given in Figure \ref{atj}.
-
-The first two rules are standard in bidirectional type systems: variables synthesize types and if a term synthesizes a type, it can be analyzed against a type (cf.
-Listing \ref{context}).
-Variables translate to variables, and if we are simply converting from synthesis to analysis, translation is not affected.
-
-The rule $\fvar{att-lam}$ must take into account the fact that, because we support type-level computation, the type annotation on the argument may not be in normal form.
+\begin{enumerate}
+\item The rule $\textsc{att-flip}$ is standard in bidirectional type systems (without subtyping): if a term synthesizes a type, it can be analyzed against that type; the translation is unaffected by this.
+\item The rule $\textsc{att-var}$ says that variables synthesize the type they have in the typing context and translate to variables in the internal language.
+\item The rule $\textsc{att-asc}$ says that a term ascribed by a type synthesizes that type if the term can be  analyzed against that type, after it has been normalized.
+The normalization judgement for type-level terms is written $\tau \Downarrow_\fvalCtx \tau'$.
+The kinding rules (not shown here) guarantee that normalization of type-level terms cannot go wrong (we will refine what precisely this means later).
+\item The rule $\textsc{att-let-syn}$ first synthesizes a type for the bound value, then adds this binding to the context and synthesizes a type for the term the binding is scoped over.
+The translation is to a function application, in the conventional manner.
+
+To do so, however, we must also translate the type itself so that an appropriate type annotation for the function argument can be emitted.
+Every type has an internal type associated with it called its \emph{representation type}.
+The type-level operator $\FF{repof}(\tau)$ evaluates to a \emph{quoted internal type}, $\titype{\sigma}$ (of kind $\kIType$), where $\sigma$ is the representation type of $\tau$.
+Type constructors define the representation type for every possible index by providing a type-level function called the \emph{representation schema}, written after the keyword $\FF{rep}$.
+The normalization rule showing this is given in Fig.
+\ref{tleval}.
+In our example, the representation schema is simple: natural numbers translate to integers.
+We will see an example where this is less trivial later.
+Note that quoted internal types support splicing using the $\dup{\tau}$ form.
+These forms are eliminated during normalization.
+\item The rule $\textsc{att-lam}$ says that lambda terms can only be analyzed against arrow types and translate to lambda terms in the internal language.
+\item The rule $\textsc{att-intro-ana}$ says that generalized introductory forms can only be analyzed against a type.
+That type constructor is consulted to extract its \emph{introductory operator definition}, written (e.g.
+in Figure \ref{nat-atfrag}) after the $\FF{iana}$ keyword as a type-level function.
+This function is given the provided operator index, the type's type index and a list encapsulating the operator's  \emph{arguments}.
+An argument is reified as a type-level term of the form $(\Gamma, e)?$ and has kind $\FF{Arg}$.
+Note that only the compiler should construct such a term (in practice, this could be enforced purely syntactically, so we do not enforce this judgmentally here).
+The operator definition is responsible for producing a translation, or evaluating to $\terr$ if this is not possible.
+A translation is simply a quoted internal term, written $\titerm{\iota}$, with kind $\kITerm$.
+Like quoted internal types, quoted internal terms support an unquote form, written $\iup{\tau}$, which is normalized away (in a capture-avoiding manner, not shown).
+
+Natural numbers have two introductory forms, each indexed trivially.
+Because there is only one generalized introductory form, we instead index the operator by a simple sum kind with two trivial cases, $\kunit + \kunit$.
+The first case corresponds to the zero operator and the second to the successor.
+In the introductory operator definition, we see in the first case that the translation $0$ is produced after checking that no arguments were provided (using a simple helper function, $\tvar{arity0}$, not shown).
+In the second case, we first extract the single argument (using the helper function $\tvar{arity1}$, not shown).
+We then analyze it against the type $\ttype{nat}{\tunit}$ using one of the elimination forms for arguments, $\FF{ana}(\tau_1; \tau_2; \tvar{t}_{trans}.\tau_3)$.
+If it succeeds, we construct the appropriate translation based on the translation of the argument.
+The normalization rule for successful analysis, $\textsc{ana}$, is shown in Figure \ref{tleval}.
+
+Note that we do not show here the error propagation rules.
+We need two additional judgements for this: the judgement $\Gamma \vdash_\fvalCtx e~\mathtt{error}$ says that $e$ has a type error, and the judgement $\tau~\mathtt{err}_\fvalCtx$ says that normalizing $\tau$ produced an $\terr$ term.
+Appropriate error propagation rules need to be written down.
+
+If analysis of the introductory form succeeds, the translation is checked for \emph{representational consistency}: that the produced translation has the type indicated by the type constructor's representation schema.
+This is expressed by essentially a standard typechecking judgement for the internal language, written $\Gamma \vdash_\fvalCtx \iota : \sigma$ (not shown).
+Note that the typing context will need to be translated to a corresponding typing context for the internal language.
+We will discuss this further below.
+
+Notice that the assisted introductory form in Figure \ref{desugaring} allows us to define type-level variables $\tvar{z}$ and $\tvar{s}$ that simultaneously provide an operator index and type ascription, simulating introductory forms that synthesize a type without requiring support for such in the semantics (we will discuss a more expressive variant of the system in Sec.
+\ref{ace} where ascriptions can also be type constructors, not just types, permitting a form of synthetic introductory form).
+
+Note also that lambda is technically the introductory form for the $\fvar{arrow}$ type constructor, but because it requires manipulating the context, it must be built in to @$\lambda$.
+We do not currently support type system fragments that require adding or manipulating existing contexts.
+
+\item The rule $\textsc{att-elim-syn}$ says that elimination forms synthesize a type.
+This is done again by consulting a type-level function, called the \emph{eliminatory operator definition}, associated with a type constructor, here the type constructor of the type  recursively synthesized for the primary operand, $e$.
+This definition is invoked with the type index, the operator index, the translation of the primary operand and a list of arguments.
+Because elimination forms synthesize a type, this definition must produce both a type and a translation.
+Representational consistency is then checked.
+
+In our example, the elimination form for natural numbers is the recursor (shown in Fig.
+\ref{natfrag}).
+Our definition of it first checks that exactly two arguments were provided (not including the primary operand), then synthesizes a type and extracts a translation from the first argument using the form $\FF{syn}(\tau_1; \tvar{t}_{ty}, \tvar{t}_{trans}.\tau_2)$ (Fig.
+\ref{tleval}).
+The second argument is analyzed against an appropriate arrow type to support variable binding.
+If successful, the type $\tvar{t1}$ (the type of the base case) is synthesized and a translation implementing the dynamic semantics of the recursor by a fixpoint computation is produced.
+
+The assisted elimination form shown in Fig.
+\ref{desugaring} provides a means to avoid providing an operator index explicitly.
+Here, we simply define $\tvar{natrec}$ as the trivial value to avoid needing to write it explicitly (and for clarity).
+
+Note that the elimination form for the $\fvar{arrow}$ type constructor can be defined in this manner (because it does not require binding variables), so application is simply syntactic sugar, rather than a built-in operator in the external language.
+\end{enumerate}
+
+%The rule $\fvar{att-lam}$ must take into account the fact that, because we support type-level computation, the type annotation on the argument may not be in normal form.
 Thus, we evaluate it to normal form.
@@
 The evaluation semantics remove quotations, so the normal form of a quoted internal type contains an internal type with no quotations.
-
-Lambda functions are the only introductory form requiring special support in the calculus (because they need to  manipulate the context; see Discussion).
+%
+%Lambda functions are the only introductory form requiring special support in the calculus (because they need to  manipulate the context; see Discussion).
 The next three rules show how any other abstractions that we define (e.g.
 records, decimals, etc.) make use of a generalized introductory form.
-
-The rule  $\fvar{att-i-unasc}$ shows that unascribed introductory forms can only be analyzed against a type.
+%
+%The rule  $\fvar{att-i-unasc}$ shows that unascribed introductory forms can only be analyzed against a type.
 Given such a type, the rule extracts a definition, named $\FF{iana}$, from the type constructor (cf.
@@
 In this example, records with two or more fields translate to nested tuples.
-
-Because we have a representation type associated with the type, we can check that the translation is \emph{representationally consistent} (the final premise).
+%
+%Because we have a representation type associated with the type, we can check that the translation is \emph{representationally consistent} (the final premise).
 As we will discuss, representational consistency combined with type safety of the internal language implies type safety overall (it arises as a strengthening of the inductive hypothesis needed to prove type safety, and is closely related to work on \emph{type-preserving compilation} in the TIL compiler for Standard ML, \cite{tarditi+:til-OLD}).
-
-The next rule, $\fvar{att-i-asc-ty}$ states that introductory forms ascribed with a type are analyzed against that type.
-
-Introductory forms ascribed with a type constructor can, according to the final introductory rule, $\fvar{att-i-asc-tycon}$, synthesize a type via the definition $\FF{isyn}$.
+%
+%The next rule, $\fvar{att-i-asc-ty}$ states that introductory forms ascribed with a type are analyzed against that type.
+%
+%Introductory forms ascribed with a type constructor can, according to the final introductory rule, $\fvar{att-i-asc-tycon}$, synthesize a type via the definition $\FF{isyn}$.
 It is passed the operator index (there is no type index, since we only have a type constructor) and a list of arguments, as before.
@@
 We check representational consistency of the result, as before.
-
-Finally, we show the rule for elimination forms.
+%
+%Finally, we show the rule for elimination forms.
 It operates by first synthesizing a type for the ``primary'' subterm, then extracting the $\FF{esyn}$ definition from its type constructor.
@@
 
-\subsection{Metatheory}
-The metatheoretic properties of interest are: type safety and termination of the type-level language (guaranteed by our kind system, though the details will need to be provided in an appendix; termination is non-trivial), type safety and decidability of the internal language (it is a standard variant of PCF, so this is trivial) and representational consistency (discussed above, follows because we check it explicitly).
+\subsection{Representational Consistency Implies Type Safety}\label{repcon}
+A key invariant that our operator definitions maintain is that well-typed external terms of type $\ttype{Nat}{\tunit}$ always translate to well-typed internal terms of internal type $\dint$, the representation type of $\ttype{Nat}{\tunit}$.
+Verifying this for the zero case is simple.
+For the translation produced by the successor case to be of internal type $\dint$ requires that $\iup{\tvar{x}}$ be of internal type $\dint$.
+Because it is the result of analyzing an argument against $\ttype{Nat}{\tunit}$, and the only other introductory form is the zero case, this holds inductively.
+We will discuss the recursor later, but it also maintains this invariant inductively.
+
+This invariant would not hold if, for example, we allowed the type and translation: $$({{\ttype{Nat}{\tunit},\titerm{(0, ())}}})$$
+\noindent
+In this case, there would be two different internal types, $\dint$ and $\dpair{\dint}{\dunit}$,  associated with a single external type, $\ttype{Nat}{\tunit}$.
+This would make it impossible to reason \emph{compositionally} about the translation of an external term of type $\ttype{Nat}{\tunit}$, so our implementation of the successor would produce ill-typed translations in some cases but not others.
+Similarly, we wouldn't be able to write functions over all natural numbers because there would not be a well-typed translation to give to such a function.
+This violates type safety: there are now well-typed external terms, according to the active typechecking and translation judgement, for which evaluating the corresponding translation would ``go wrong''.
+
+To reason compositionally about the semantics of well-typed external terms when they are given meaning by translation to a typed internal language, the system must maintain the following property: for every  type, $\tau$, there must exist an internal type, $\sigma$, called its \emph{representation type}, such that the translation of every external term of type $\tau$ has internal type $\sigma$.
+This principle of \emph{representational consistency} arises essentially as a strengthening of the inductive hypothesis necessary to prove that all well-typed external terms translate to well-typed internal terms because $\lambda$ and successor are defined compositionally.
+It is closely related to the concept of \emph{type-preserving compilation} developed by Morrisett et al.
+for the TIL compiler for SML \cite{tarditi+:til-OLD}, here lifted into the language.
+Our judgements check this extension correctness property directly by typechecking each translation produced by a user extension (the translations of, for example, lambda terms will be inductively representationally consistent, so no additional check is needed).%That is, instead of being used as a necessary condition for compiler correctness, we are using it as a sufficient condition for type safety.
+%We emphasize the distinction between translation (which in our calculus endows terms with a dynamic semantics) and compilation (which must preserve the dynamic semantics of terms).
+%In the presence of extensions, checking for representational consistency becomes subtle if we wish to guarantee conservativity, as we will discuss in the next subsection.
+
+%For the semantics to ensure that representational consistency is maintained by all operator definitions, we require, as briefly mentioned in Sec.
+\ref{types}, that each type constructor declare a \emph{representation schema} after the keyword \textsf{schema}.
+This is a type-level function of kind $\karrow{\kappaidx}{\kIType}$, where $\kappaidx$ is the index kind of the type constructor.
+The kind $\kIType$ has introductory form $\titype{\bar \sigma}$ and no elimination form, and is similar to the kind $\kITerm$ introduced above.
+There are two forms of \emph{translational internal type}, $\bar \sigma$, that do not correspond to forms in $\sigma$.
+As with translational internal terms, these are included to allow an internal type to be formed compositionally:
+%\begin{enumerate}
+%\item $\dup{\tau}$ splices in another translational internal type $\tau$
+%\item $\trepof{\tau}$ refers to the representation type of type $\tau$
+%\end{enumerate}
+%
+%When the semantics (i.e.
+the compiler) needs to determine the representation type of the type $\ttype{Tycon}{\tauidx}$ it simply applies the representation schema of $\fvar{Tycon}$ to ${\tauidx}$.
+Note that the representation type of a type cannot be extracted directly from within the type-level language, again for reasons that we will discuss below.
+%
+%These additional forms are not needed by the representation schema of $\fvar{Nat}$  because it is trivially indexed.
+In Fig.
+\ref{tuple}, we will discuss an example of the type constructor $\fvar{Tuple}$, implementing the semantics of $n$-tuples by translation to nested binary products.
+Here, the representation schema refers to the representations of the tuple's constituent types and is computed by folding over a list, so both of these are used\todo{add this example from appendix of ESOP}.
+
+Translational internal terms can contain translational internal types.
+We see this in the definition of the recursor on natural numbers.
+The type assignment is the arbitrary type, $\tvar{t2}$.
+We cannot know what the representation type of $\tvar{t2}$ is, so we refer to it abstractly using $\trepof{\tvar{t2}}$.
+Luckily, the proof of representational consistency of this operator is parametric over the representation type of $\tvar{t2}$.
+
+\subsection{Metatheory}\label{theory}
+%\newcommand{\F}[1]{\textsf{#1}~}
+%\newcommand{\FF}[1]{\textsf{#1}}
+%\newcommand{\Q}{\FF{Arg}}
+%\renewcommand{\tnil}[1]{[]}
+%\begin{figure}[t]
+%\small
+%$$\begin{array}{rcl}	
+%%\textbf{programs} & 
+%\rho & ::= & \F{tycon}\fvar{tycon}~\F{of}\kappaidx~{\{}TC{\}}; \rho \pipe e\\ %\pfam{\familyDf}{\progsort}
+%%\\&  \pipe & \pdef{t}{\kappa}{\tau}{\progsort}  \pipe e
+%%\pipe \pdef{t}{\kappa}{\tau}{\progsort} 
+% TC & ::= & \F{iana} \{\tau\}; \F{isyn} \{\tau\}; \F{esyn} \{\tau\}; \F{rep} \{\tau\}\\
+%\\ 
+%e & ::= & \evar{x} \pipe \elam{\evar{x}}{\tau}{e} \pipe I \pipe I : \tau \pipe I :: \fvar{tycon} \pipe E\\
+%I & ::= &  \FF{intro}[\tauidx](\splat{e}{1}{n})\\
+%E & ::= & e\cdot\FF{elim}[\tauidx](\splat{e}{1}{n})\\
+%\\
+% \tau 	& ::= 	& 	\tvar{t} \pipe 
+%														\tlam{t}{\kappa}{\tau} \pipe 
+%														\tapp{\tau_1}{\tau_2}  
+%\\&\pipe&											
+%														\tnil{\kappa} \pipe \tcons{\tau_1}{\tau_2} \pipe 
+%									                     \tfold{\tau_1}{\tau_2}{h}{t}{r}{\tau_3}
+%														\\
+% 		& \pipe	& 	  \ell \pipe 
+%		\tunit \pipe 
+%														\tpair{\tau_{1}}{\tau_{2}} \pipe 
+%														\tfst{\tau} \pipe 
+%														\tsnd{\tau} \pipe \cdots 
+%														\\	
+%%     & \pipe & \tinl{\kappa_2}{\tau_1} \pipe \tinr{\kappa_1}{\tau_2} \\
+%%     &\pipe& \tsumcase{\tau}{t}{\tau_1}{t}{\tau_2} \\
+%&\pipe	& 	\ttype{tycon}{\tauidx} \\&\pipe& \tfamcase{\tau}{tycon}{x}{\tau_1}{\tau_2}\\
+%&\pipe & 					\tifeq{\tau_{1}}{\tau_{2}}{\kappa}{\tau_{3}}{\tau_{4}} 
+%%														\\													%				& & \pipe & \tfamcase{\tau}{Fam}{x}{\tau_1}{\tau_2}\\
+%%																								
+%\\ 		 & 	\pipe	&	\tden{\tau_2}{\tau_1} \pipe \terr %\pipe \tden{\ibar}{\tau}^{\checkmark} 
+%\pipe \ttypeof{\tau} \pipe \itransof{\tau} \\% \tdencase{\tau}{x}{t}{\tau_1}{\tau_2}\\
+%% %& & \pipe & 
+%%%														\tdencase{\tau}{y}{x}{\tau_1}{\tau_2}
+%%%														 \\
+%%
+% & \pipe & \titerm{\iota}
+%			\pipe	\titype{\sigma} \pipe \trepof{\tau} \\
+%		& \pipe &  (\Gamma; e)? \pipe \FF{syn}(\tau) \pipe \FF{ana}(\tau; \tau')\\
+%%\textbf{external terms} 				&	e	&	::=	&	\evar{x} \pipe 
+%%%														\efix{x}{\tau}{e} \pipe 
+%%														\elam{\evar{x}}{\tau}{e} \pipe 
+%%														\eop{tycon}{op}{
+%%															\tauidx
+%%														}{
+%%  												    		\splat{e}{1}{n}
+%%														} \\
+%%									& 		&		& 	\\
+%%
+%%
+%%												
+%%%\text{deabstracted}& \iota & ::= & \mathcal{G}[\iota, \sigma]\\
+%%												
+%%\\
+%%\\
+%%							
+%%% &  & \pipe & \tvalof{\tau_1}{\tau_2} \pipe \iup{\tau} \\
+%%% 												\trepof{\tau} \pipe \dup{\tau}\\
+%%\text{translational IL}	& \bar{\iota} & ::= & x \pipe \ifix{x}{\bar \sigma}{\bar \iota} 
+%%	%\pipe \ilam{x}{\bar \sigma}{\bar \iota} \pipe \iapp{\bar \iota_1}{\bar \iota_2} 
+%%	\pipe \cdots \pipe \itransof{\tau} \\						
+%% & \bar{\sigma} & ::= & \darrow{\bar \sigma_1}{\bar \sigma_2} \pipe \cdots \pipe \dup{\tau} \pipe \trepof{\tau} \\
+%%%\text{abstracted} & \sabs & ::= & \darrow{\sabs_1}{\sabs_2} \pipe \cdots \pipe \sabsrep{\tau}
+%%											\\
+%\kappa	&	::=	&	\karrow{\kappa_1}{\kappa_2} \pipe \klist{\kappa} \pipe \FF{L} \pipe \kunit \pipe \kpair{\kappa_1}{\kappa_2} \pipe \cdots \\
+% & \pipe & \kTypeBlur \pipe \FF{TT} \pipe \kITerm \pipe \kIType \pipe \Q
+%%											    \klabel \pipe
+%%											    \klist{\kappa} \pipe
+%%												\kunit \pipe 
+%%												\kpair{\kappa_{1}}{\kappa_{2}} \\
+%%												&&\pipe&
+%%												\ksum{\kappa_1}{\kappa_2} \pipe
+%%												\kTypeBlur \pipe \kDen \pipe 
+%%												\kIType								
+%%%\textbf{ops signature}			& \Theta	&	::=	&	\kOpEmpty \pipe \kOp{\Theta}{op}{\kappai}\\
+%%%											 							&		&		&	\\
+%\\
+%\\
+%\iota	&	::=	&	\evar{x} \pipe 
+%												\ifix{\evar{x}}{\sigma}{\iota} \pipe
+%												\ilam{\evar{x}}{\sigma}{\iota} \pipe 
+%												\iapp{\iota_{1}}{\iota_{2}} 
+%												\\ & \pipe & \iunit \pipe \ipair{\iota_1}{\iota_2} \pipe \ifst{\iota} \pipe \isnd{\iota} \pipe \cdots \pipe \iup{\tau}
+%%\\&\pipe&								\cdots \pipe 	
+%%												n \pipe \iop{\iota_{1}}{\iota_{2}} \pipe \iIfEq{\iota_{1}}{\iota_{2}}{\dint}{\iota_{3}}{\iota_{4}} \\
+%%& \pipe & 
+%%												\iunit \pipe
+%%												\ipair{\iota_{1}}{\iota_{2}} \pipe 
+%%												\ifst{\iota} \pipe
+%%												\isnd{\iota} 
+%%\\&\pipe&												
+%%												 \iinl{\sigma_2}{\iota_1} \pipe \iinr{\sigma_1}{\iota_2} \\&\pipe & \icase{e}{x}{e_1}{x}{e_2} \\
+%\\	\sigma	&	::=	&    \darrow{\sigma_1}{\sigma_2} \pipe \dunit \pipe \dpair{\sigma_1}{\sigma_2} %\pipe \dint \pipe \dunit \pipe \dpair{\sigma_1}{\sigma_2} \pipe \dsum{\sigma_1}{\sigma_2}  
+%\pipe \cdots \pipe \dup{\tau}
+%\end{array}$$
+%%\vspace{-10pt}
+%\caption{\small Syntax of Core $\lamAce$.
+Here, $x$ ranges over external and internal language variables, $\tvar{t}$ ranges over type-level variables, $\fvar{tycon}$ ranges over type constructor names and $\ell$ ranges over labels.
+%\label{grammar}}
+%\end{figure}
+%\begin{figure}
+%\small
+%\begin{flalign}
+%& \F{tycon}\fvar{record}~\F{of}\klist{\kpair{\FF{L}}{\kTypeBlur}}~\{\\
+%& \quad \FF{iana}~\{\tlam{i}{\kpair{\klist{\kpair{\FF{L}}{\kTypeBlur}}}{\klist{\FF{L}}}}{\tlam{a}{\klist{\Q}}{\\& \quad\quad\tfold{\tvar{zip3}~\tfst{\tvar{i}}~\tsnd{\tvar{i}}~\tvar{a}}
+%	 {\titerm{\tunit}}{h}{t}{r}{
+%	\\&\quad\quad\quad\tifeq{\tfst{\tvar{first}~\tvar{h}}}{\tvar{second}~\tvar{h}}{\FF{L}}{
+%	\\&\quad\quad\quad\quad
+%	\F{let}~\tvar{x}:\kITerm~=~\FF{ana}(\tvar{third}~\tvar{h}; \tsnd{\tvar{first}~\tvar{h}})
+%~\FF{in}
+%\\ & \quad\quad\quad\quad \tfold{\tvar{t}}{\tvar{x}
+%}{\_}{\_}{\_}{\titerm{(\iup{\tvar{x}},\iup{\tvar{r}})}}	}{\terr}
+%%     \F{let}~\tvar{tt}:\FF{TT}=~\FF{syn}(\tsnd{\tsnd{\tvar{h}}}) \F{in}\\
+%%\\     & \quad\quad\quad 
+%}
+%}}\}\\
+%& \quad \FF{isyn}~\{\tlam{i}{\klist{\FF{L}}}{\tlam{a}{\klist{\Q}}{
+%\\&\quad\quad \tfold{\tvar{zip2}~\tvar{i}~\tvar{a}}{\tden{\titerm{()}}{\ttype{record}{\tnil{{\kpair{\FF{L}}{\kTypeBlur}}}}}}{h}{t}{r}{
+%\\&\quad\quad\quad \F{let}~\tvar{htt}:\FF{TT}~=~\FF{syn}(\tsnd{\tvar{h}})
+%~\FF{in}
+%\\&\quad\quad\quad \F{let}~\tvar{hty}:\kTypeBlur~=~\ttypeof{\tvar{htt}}
+%\\&\quad\quad\quad \F{let}~\tvar{ty}:\kTypeBlur~=~\ttype{record}{(\tfst{\tvar{h}}, \tvar{hty})::\tnil{{\kpair{\FF{L}}{\kTypeBlur}}}}
+%~\FF{in}
+%\\&\quad\quad\quad \tfold{\tvar t}{\tden{\itransof{\tvar{htt}}}{\tvar{ty}}}{\_}{\_}{\_}{
+%\\&\quad\quad\quad\quad \tden{\titerm{(\iup{\itransof{\tvar{htt}}},\iup{\itransof{\tvar{r}}})}}{\tvar{ty}}}}
+%}}\}\\
+%& \quad \FF{esyn}~\{\tlam{i}{\FF{L}}{\tlam{a}{\klist{\Q}}{
+%	\tvar{arity1}~\tvar{a}~\tlam{ty}{\kTypeBlur}{\tlam{x}{\kITerm}{
+%\\&\quad\quad		\tfamcase{\tvar{ty}}{record}{sig}{
+%\\&\quad\quad\quad \tfold{\tvar{sig}}{\terr}{h}{t}{r}{
+%\\&\quad\quad\quad\quad \tifeq{\tfst{\tvar{h}}}{\tvar{i}}{\FF{L}}{\tfold{\tvar{t}}{\tden{\tvar{x}}{\tsnd{\tvar{h}}}}{\_}{t'}{\_}{
+%\\&\quad\quad\quad\quad\quad \tfold{\tvar{t'}}{\tden{\titerm{\ifst{\iup{\tvar{x}}}}}{\tsnd{\tvar{h}}}}{\_}{\_}{\tvar{r'}}{
+%\\&\quad\quad\quad\quad\quad\quad \tden{\titerm{\isnd{\iup{{\itransof{\tvar r}}}}}}{\tsnd{\tvar{h}}}}}\\&\quad\quad\quad\quad\hspace{-3px}}{\tvar{r}}
+%}\\&\quad\quad\hspace{-3px}}{\terr}
+%	}}
+%}}\\
+%& \quad \FF{rep}~\{\tlam{i}{\klist{\kpair{\FF{L}}{\kTypeBlur}}}{ 	\\& \quad \quad \tfold{\tvar{i}}{\titype{\dunit}}{s}{j}{r}{
+% 		\tfold{\tvar{j}}{\trepof{\tsnd{\tvar{s}}}}{\_}{\_}{\_}{\\
+%		&\quad\quad\quad
+% 		\titype{\dpair{\dup{\trepof{\tsnd{\tvar{s}}}}}{\dup{\tvar{r}}}}
+% 		}
+% 	}}\}\\
+%& \}; \\
+%& \F{let}\tvar{R} : \kTypeBlur = \ttype{record}{(\ell_1, \ttype{record}{\tnil{{\FF{L}\times\kTypeBlur}}}) :: \tnil{{\FF{L}\times\kTypeBlur}}}~\F{in}\\
+%&\F{let}id = \elam{x}{\tvar{R}}{\{\ell_1=~x\cdot\ell_1\} :: \fvar{record}}\\
+%&\F{let}triv = id[\{\ell_1 = \{\}\}]\cdot\ell_1
+%%&\F{let}one = \FF{intro}[\ell_\text{succ}]
+%%&\F{let}plus = \elam{x}{\tvar{nat}}{\elam{y}{\tvar{nat}}{\FF{elim}[()](x; y; \\
+%%& \quad \elam{r}{\tvar{nat}}{intro[\ell_\text{succ}](r) : \tvar{nat}}})}
+%\end{flalign}
+%\caption{The definition of the record type in $\lamAce$ using the desugarings in Figure \ref{desugaring} and with the addition of simple let bindings for both type and term variables (not shown).
+Some type-level helper functions also omitted for concision.}
+%
+%\label{record-theory}
+%\end{figure}
+%We will now give a core typed lambda calculus, $\lamAce$, that captures the semantics described in the previous sections.
+It is intended to make precise how active typechecking and translation works and how our mechanism relates to existing work on bidirectional typechecking, type-level computation and typed compilation while abstracting away from the details of Python's syntax and imposing a stronger type-level semantics that will allow us to state metatheoretic properties of interest.
+We will assume a fixed base for functions (providing the standard semantics of lambda functions) and target language, which we here call the \emph{internal language}.
+%
+%The syntax of $\lamAce$ is shown in Figure \ref{grammar} and an example \emph{program} that defines a type constructor, $\fvar{record}$, with a semantics similar to that given in Listing \ref{record}\footnote{Technically, this defines labeled tuples, because the order of labels matters.}, using it to write an identity function and compute the value $triv$ (the empty record), is shown in Figure \ref{record-theory}.
+%
+%%In this section, we will develop an ``actively typed'' version of the simply-typed lambda calculus with simply-kinded type-level computation called $\lamAce$.
+More specifically, the level of types, $\tau$, will itself form a simply-typed lambda calculus.
+\emph{Kinds} classify type-level terms in the same way that types conventionally classify expressions.
+Types become just one kind  of type-level value (which we will write $\kTypeBlur$, though it is also variously written $\star$, \verb|T| and \verb|Type| in various settings).
+Rather than there being a fixed set of type and operator constructors, we allow the programmer to declare new  constructors, and give their static and dynamic semantics by writing type-level functions.
+The kind system combined with techniques borrowed from the typed compilation literature and a form of type abstraction will allow us to prove strong type safety, decidability and conservativity theorems.
+%%
+%%The syntax of Core $\lamAce$ is given in Fig.
+\ref{grammar}.
+An example of a program defining type and operator constructors that can be used to construct an active embedding of G\"odel's \textbf{T} into $\lamAce$ is given in Fig.
+\ref{nat}.
+We will discuss its semantics and how precisely the embedding, seen being used starting on line 15 to ultimately compute the sum of two and two, works as we go on.
+Natural numbers can, of course, be isomorphically embedded in existing languages, with a similar usage and asymptotic performance profile (up to function call overhead as an abstract type, for example).
+We will provide more sophisticated examples where this is less feasible later on (and note that type abstraction is an orthogonal mechanism).
+%%a type constructor declaration, $\fvar{Nat}$, indexed trivially, together with three operator constructors, also all indexed trivially, that implement the standard introductory forms for natural numbers as well as the recursor operator (as in G\"odel's T \cite{pfpl}).
+Following the type constructor declaration, we apply $\fvar{Nat}$ with the trivial index, $\tunit$, to form the type $\tvar{nat}$.
+Finally, we write an external term that uses the operators associated with $\fvar{nat}$ and the built-in constructor $\fvar{Parr}$, governing partial functions, to define an addition function and compute the addition of the natural numbers  two and two.
+We will introduce a more convenient concrete syntax in later portions of this thesis; for now we will restrict ourselves to the abstract syntax so that this example can directly aid in understanding the semantics.
+%
+%\subsection{Overview}\label{programs}
+%
+%A \emph{program}, $\rho$, consists of a series of constructor declarations followed by an external term, $e$.
+Compiling a program consists of first \emph{kind checking} it (see below), then typechecking the external term and simultaneously {translating} it to a term, $\iota$, in the {typed internal language}.
+%%These correspond to the premises of the \emph{central compilation judgement} $\pkcompiles{\rho}{\iota}$:
+%%\[
+%%\inferrule[p-compiles]{
+%%	\emptyset \vdash_{\fvalCtx_0} \rho\\
+%%%	\progOK{\emptyset}{\fvalCtx_0}{\rho}\\
+%%	\pcompiles{\fvalCtx_0}{\rho}{\iota}
+%%}{\pkcompiles{\rho}{\iota}}
+%%\]
+%%This anchors our exposition; we will describe how it is derived (i.e. how to write a compiler for $\lamAce$) in the following sections. 
+%The key judgements are the \emph{bidirectional active typing judgements}  (Fig.
+\ref{att}, which we describe starting in Sec.
+\ref{opcons}).
+They relate an external term, $e$, to a {type}, $\tau$, called its \emph{type assignment}, and an internal term, $\iota$, called its \emph{translation}, under \emph{typing context} $\Gamma$ and \emph{constructor context} $\Phi$.
+The first is synthesis, the second analysis.
+%\[\Gamma \vdash_\fvalCtx e \Rightarrow \tau \leadsto \iota
+%~~~~~~~
+%\Gamma \vdash_\fvalCtx e \Leftarrow \tau \leadsto \iota\]
+%%\[\ecompilesAX{e}{\tau}{\iota}\]
+%
+%The typing context, $\Gamma$, maps variables to types in essentially the conventional way (\cite{pfpl} contains the necessary background for this section).
+The constructor context, $\Phi$, tracks user-defined type  constructors.
+% Weakening and exchange of the constructor context closely related to the issue of conservativity that we will return to.
+Each constructor is identified by name, so there is no analog to contraction.
+%
+%The dynamic behavior of an external term is determined entirely by its translation to the internal language, which has a conventional operational semantics.
+This form of semantics can be seen as lifting into the language specification the first stage of a type-directed compiler like the TIL compiler for Standard ML \cite{tarditi+:til-OLD} and has some parallels to the Harper-Stone semantics for Standard ML, where external terms were also given meaning by elaboration from the EL to an IL \cite{Harper00atype-theoretic}.
+%
+%
+%In $\lamAce$, the internal language (IL) provides partial functions (via the generic fixpoint operator of Plotkin's PCF) and simple product  types for the sake of our example.
+In practice, the internal language could be any typed  language with a specification for which type safety and decidability of typechecking have been satisfyingly determined.
+The internal type system serves as a ``floor'': guarantees that must hold for terms of any type (e.g.
+that out-of-bounds access to memory never occurs) must be maintained by the internal type system.
+User-defined constructors can enforce invariants stronger than those the internal type system maintains at particular types, however.
+Performance is also ultimately limited by the internal language and downstream compilation stages that we do not here consider (safe compiler extension has been discussed in previous work, e.g.
+\cite{conf/pldi/TatlockL10}).
+%
+%
+%\subsection{Types and Type-Level Computation}\label{types}
+%$\lamAce$ supports, and makes extensive use of, simply-kinded type-level computation.
+Specifically, type-level terms, $\tau$, themselves form a typed lambda calculus.
+The classifiers of type-level terms are called \emph{kinds}, $\kappa$, to distinguish them from  \emph{types}.
+Types are  type-level values of kind $\kTypeBlur$.
+In Ace, the type-level language (together with the level of programs) is written in Python, which can be thought of as having a rather limited kind system (with one kind, \verb|dyn|).
+Here, we are able to more precisely discuss the kinds of values in the type-level language.
+%
+%In most languages, types are formed by applying one of a collection of \emph{type constructors} to zero or more \emph{indices}.
+In $\lamAce$, the situation is notionally similar.
+User-defined type constructors can be declared at the top of a program (or lifted to the top, in practice) using \textsf{tycon}.
+Each constructor in the program must have a unique name, written e.g.
+\fvar{record}.\footnote{We assume naming conflicts can be avoided by some extrinsic mechanism.} A type constructor must also declare an \emph{index kind}, $\kappaidx$.
+%
+% To permit the embedding of interesting type systems, the type-level language includes several kinds other than $\kTypeBlur$.
+We lift several functional data structures to the type level: here, only unit ($\kunit$), binary products ($\kpair{\kappa_1}{\kappa_2}$) and lists ($\klist{\kappa}$), in addition to labels (introduced as $\ell$, possibly with a subscript, having kind \FF{L}).
+Our record type constructor is indexed by a list of pairs of {labels}  and types (a signature, in essence; line 1).
+The type constructor $\fvar{arrow}$ is included in the initial constructor context, $\fvalCtx_0$ and has index kind $\kpair{\kTypeBlur}{\kTypeBlur}$.
+% In practice, one could include a richer functional programming language and retain the spirit of the calculus, as long as it does not introduce general recursion at the type level. 
+%  
+%A type is introduced by applying a type constructor to an index of this kind, written $\ttype{Tycon}{\tauidx}$.
+For example, the type of natural numbers is indexed trivially, so it is written $\ttype{Nat}{\tunit}$.
+We see a record type, abbreviated $\tvar{R}$, constructed on line 26.
+%
+%The kind $\kTypeBlur$ also has an elimination form, $$\tfamcase{\tau}{Tycon}{x}{\tau_1}{\tau_2}$$ allowing the extraction of a type index by case analysis against a contextually-available type constructor.
+To a first approximation, one might think of type constructors as constructors of a built-in open \cite{conf/ppdp/LohH06}, $\kTypeBlur$, at the type-level.
+Like open datatypes, there is no notion of exhaustiveness so the default case is required for totality.
+We will see where this is used shortly.
+%
+%
+%
+%%We will write the kind of types as $\kTypeBlur$, though it is also written $\star$ or \verb|Type| in various similarly structured languages (see Sec. \ref{related-work}). 
+%% Rather than there being a fixed set of type constructors, we allow the programmer to declare new type  constructors, and give the static and dynamic semantics of their associated operators, by writing type-level functions.
+In the semantics for this calculus, our kind system combined with techniques borrowed from the typed compilation literature and a form of type abstraction allow us to prove strong type safety, decidability and conservativity theorems.
+%
+%
+%
+%% and integers ($\dint$).
+We also include labels ($\klabel$), written in a slanted font, e.g.
+$\tlabel{myLabel}$, which are string-like values that only support comparison and play a distinguished role in the expanded syntax, as we will later discuss.
+Our first example, $\fvar{nat}$, is indexed trivially, i.e.
+by unit kind, $\kunit$, so there is only one natural number type, $\ttype{nat}{\tunit}$, but we will show examples of type constructors that are indexed in more interesting ways in later portions of this work.
+For example, $\fvar{LabeledTuple}$ has index kind $\klist{\kpair{\klabel}{\kTypeBlur}}$.
+% 
+% Type constructors are not first-class; they do not themselves have arrow kind as in some kind systems (e.g.
+\cite{watkins2008specifying}; Ch.
+22 of \emph{PFPL} describes a related system \cite{pfpl}).
+The type-level language does, however, include total functions of arrow kind, written $\karrow{\kappa_1}{\kappa_2}$.
+Type constructor application can be wrapped in a type-level function to emulate a first-class or uncurried version of a type constructor for convenience (indeed, such a wrapper could be generated automatically, though we do not do so).
+%
+%Two type-level terms of kind $\kTypeBlur$ are equivalent if they apply the same constructor, identified by name, to equivalent indices.
+Going further, we ensure that deciding type equivalence requires only checking for syntactic equality after normalization by imposing the restriction that equivalence at a type constructor's index kind must be decidable in this way.
+Our treatment of equivalence in the type-level language is thus quite similar to the treatment of term-level equality using ``equality types'' in a language like Standard ML.
+%% A kind $\kappa$ is an  \emph{equality kind} if $\kEq{\kappa}$ can be derived (see appendix). 
+%Conditional branching on the basis of equality at an equality kind can be performed in the type-level language.
+Equivalence at arrow kind is not decidable by our criteria, so type-level functions cannot appear within type indices.
+This also prevents general recursion from arising at the type level.
+Without this restriction, a type-level function taking a type as an argument could ``smuggle in'' a self reference as a type index, extracting it via case analysis (continuing our analogy to open datatypes, this is closely related to the positivity condition for inductive datatypes in total functional languages like Coq).% as maintaining the metatheoretic guarantee that typing respects type equivalence would impose a substantial burden in such a setting.% (a na\"ive approach to this would impose non-trivial extrinsic proof obligations onto extension developers that, unlike in others in this thesis, could threaten type safety).
+%
+%Every type constructor also defines a \emph{representation}, a type-level function that associates with every type an internal \emph{representation type} (analagous to \verb|trans_type| above).
+We will return to this after introducing the external forms and operator definitions.
+%
+%\subsection{Core External Forms and Desugaring}\label{opcons}
+%The syntax for external terms (Figure \ref{grammar}) contains variables, $\lambda$ terms, three generalized introductory forms and a single generalized elimination form.
+The introductory forms are either unascribed, ascribed with a type or ascribed with a type constructor, as in our discussion of Ace.
+These generalized forms take a single a type-level value as an {index} and $n \geq 0$ arguments, which are other external terms.
+To better motivate this choice, we can give a purely syntactic desugaring of a Python-like syntax with labels to these forms, shown in Figure \ref{desugaring}.
+It is instructive to rewrite lines 27-28 of Figure \ref{record-theory} using these desugarings:% In Ace, desugarings can be user-defined
+%\noindent
+%\vspace{-5px}
+%$$
+%\begin{array}{l}
+%\FF{let}~id=\lambda x:\tvar{R}.\FF{intro}[\ell_1 :: []](x\cdot\FF{elim}[\ell_1]()) :: \fvar{record}\\
+%\FF{let}~triv=id\cdot\FF{elim}[()](\FF{intro}[\ell_1 :: []](\FF{intro}[[]]()))\cdot\FF{elim}[\ell_1]()
+%\end{array}
+%$$
+%\vspace{-5px}
+%\begin{figure}[t]
+%\small
+%\[
+%\begin{array}{rcl}
+%\{ \ell_1{=}~e_1, \ldots, \ell_n{=}~e_n \} & := & \FF{intro}[\ell_1 :: \ldots :: \ell_n :: []](e_1; \ldots; e_n)\\
+%(e_1, ..., e_n) & := & \FF{intro}[()](e_1; \ldots; e_n)\\
+%\#n & := & \FF{intro}[n]()\\
+%"s" & := & \FF{intro}[s]()\\
+%e.\ell & := & e\cdot\FF{elim}[\ell]()\\
+%e[e_1; \ldots; e_n] & := & e\cdot\FF{elim}[()](e_1; \ldots; e_n)\\
+%e.\ell(e_1; ...; e_n) & := & e\cdot\FF{elim}[\ell](e_1; \ldots; e_n)
+%%(e_1, ..., e_n) & := & intro[()](e_1; ...; e_n)
+%%n &  := & intro[n]()
+%%s & := & intro[s]()
+%%
+%%e.l & := & elim[l](e)
+%%e[e_1; ...; e_n] & := & elim[()](e_1, ..., e_n)
+%%e.l(e_1; ...; e_n) & := & elim[l]
+%\end{array}
+%\]
+%\caption{Desugaring from conventional syntax to core forms. We assume that the type-level language has numbers, $n$, and strings, $s$ (not shown for concision).}
+%\label{desugaring}
+%\end{figure}
+%
+%%User-defined operator constructors are declared using \textsf{opcon}.
+For reasons that we will discuss, our calculus associates every operator  constructor with a type constructor.
+The \emph{fully-qualified name} of every operator constructor, e.g.
+$\fvar{Nat}.\opvar{z}$, must be unique.
+Operator constructors, like type constructors, declare an index kind, $\kappaidx$.
+In our first example, all the operator constructors are indexed trivially (by index kind $\kunit$), but other examples use more interesting indices.
+For example, in SML, the projection operator \verb|#3| can be applied to an $n$-tuple, $e$, iff $n \geq 3$.
+Note that it thus cannot be a function with a standard arrow type.
+Notionally, \verb|#| is an operator constructor and \verb|3| is its index.
+In an active embedding of $n$-tuples into $\lamAce$, this would be written $\eop{Tuple}{prj}{3}{e}$ (we will nearly recover ML's syntax later).
+$\fvar{LabeledTuple}.\opvar{prj}$ is the operator constructor used to access a field of a labeled tuple, so it has index kind $\klabel$.
+An operator itself is, notionally, selected by indexing an operator constructor, e.g.
+$\fvar{Nat}.\opvar{s}\langle \tunit \rangle$, but technically neither operator constructors nor operators are first-class at any level (additional machinery would be needed, e.g.
+an \textsf{Op} kind, but this is not fundamental to our calculus).
+%%Instead, in the external language, an operator constructor is applied by simultaneously providing an index and  $n \geq 0$ \emph{arguments}, written $\eop{Tycon}{op}{\tauidx}{\splat{e}{1}{n}}$\footnote{It may be helpful to distinguish between type/operator constructors and \emph{term formers}.
+There are term formers at all levels in the calculus.
+For example, operator constructor application and $\lambda$ are  external term formers, and type constructor application is a type-level term former.
+We might write these following Harper's conventions for abstract syntax to highlight this distinction \cite{pfpl}: $\mathtt{lam}[\tau](x.e)$, $\mathtt{ocapp}[\fvar{Tycon}, \opvar{op}, \tauidx](\splat{e}{1}{n})$ and $\texttt{tcapp}[\fvar{Tycon}](\tau)$.}.
+For example, on line 18 of Fig.
+\ref{nat}, we see the operator constructors $\fvar{Nat}.\opvar{z}$ and $\fvar{Nat}.\opvar{s}$ being applied to compute $two$.
+%\footnote{Although our focus here is entirely on semantics, a brief note on syntax: in the expanded syntax, the trivial indices and empty argument lists can be omitted, so we could write \texttt{Nat.s(Nat.s(Nat.z))}.
+With the ability to ``open'' a type's operators into the context, we could shorten this still to \texttt{s(s(z))}.
+Alternatively, with the ability to define a TSL in a manner similar to that in Sec.
+\ref{aparsing}, we might instead just write \texttt{2}.}
+%
+%\subsection{Operator Definitions and Representational Consistency}
+%
+%%
+%\newcommand{\atjsynX}[3]{\Gamma \vdash_\fvalCtx #1 \Rightarrow #2 \leadsto #3}
+%\newcommand{\atjanaX}[3]{\Gamma \vdash_\fvalCtx #1 \Leftarrow #2 \leadsto #3}
+%\begin{figure}[t]
+%\small
+%$\fbox{\inferrule{}{\ecompilesAX{e}{\tau}{\iota}}}$
+%~~~~$\Gamma ::= \emptyset \pipe \Gamma, x \Rightarrow \tau$
+%\begin{mathpar}
+%\inferrule[att-var]{
+%	x \Rightarrow \tau \in \Gamma
+%}{
+%	\atjsynX{x}{\tau}{x}
+%}
+%
+%\inferrule[att-syn-to-ana]{
+%	\atjsynX{e}{\tau}{\iota}
+%}{
+%	\atjanaX{e}{\tau}{\iota}
+%}
+%
+%\inferrule[att-lam]{
+%	\tau_1 \Downarrow_\fvalCtx \tau_1'\\
+%		\trepof{\tau_1'} \Downarrow_\fvalCtx \titype{\sigma}\\\\
+%	\Gamma, x \Rightarrow \tau_1' \vdash_\fvalCtx e \Rightarrow \tau_2 \leadsto \iota
+%%	\iota \hookrightarrow_\fvalCtx \iota'\\
+%%		\sigma \hookrightarrow_\fvalCtx \sigma'
+%%	\ddbar{\fvar{Arrow}}{\fvalCtx}{\trepof{\tau_1'}}{\sbar_1}\\
+%	%\delfromtau{$\Xi_0$}{\fvalCtx}{\tau_1'}{\sabs}\\\\
+%}{
+%	\atjsynX{\elam{x}{\tau_1}{e}}{\ttype{arrow}{(\tau_1', \tau_2)}}{\ilam{x}{\sigma'}{\iota}}
+%}
+%
+%\inferrule[att-i-unasc]{
+%	\vdash_\fvalCtx \FF{iana}(\fvar{tycon})=\taudef\\
+%	\taudef~(\tauidx, \tauidx')~((\Gamma; e_1)? :: \ldots :: (\Gamma; e_n)? :: []) \Downarrow_\fvalCtx \titerm{\iota}\\
+%%	\trepof{\ttype{tycon}{\tauidx'}} \Downarrow_\fvalCtx \titype{\sigma}\\
+%		\trepof{\ttype{tycon}{\tauidx'}} \Downarrow_\fvalCtx \titype{\sigma}\\
+%		\Gamma \vdash_\fvalCtx \iota : \sigma
+%}{
+%	\atjanaX{\FF{intro}[\tauidx](e_1; \ldots; e_n)}{\ttype{tycon}{\tauidx'}}{\iota}
+%}
+%
+%\inferrule[att-i-asc-ty]{
+%	\atjanaX{I}{\tau}{\iota}
+%}{
+%	\atjsynX{I : \tau}{\tau}{\iota}
+%}
+%
+%\inferrule[att-i-asc-tycon]{
+%	\vdash_\fvalCtx \FF{isyn}(\fvar{tycon})=\taudef\\\\
+%	\taudef~\tauidx~((\Gamma; e_1)? {::}{\ldots}{::}(\Gamma; e_n)? {::} []) \Downarrow_\fvalCtx \tden{\titerm{\iota}}{\ttype{tycon}{\tauidx'}}\\
+%			\trepof{\tau_1'} \Downarrow_\fvalCtx \titype{\sigma}\\
+%	\Gamma \vdash_\fvalCtx \iota : \sigma
+%}{
+%	\atjsynX{\FF{intro}[\tauidx](e_1; \ldots; e_n)] :: \fvar{tycon}}{\tau}{\iota'}
+%}
+%
+%\inferrule[att-elim]{
+%	\atjsynX{e}{\ttype{tycon}{-}}{-}\\
+%	\vdash_\fvalCtx \FF{esyn}(\fvar{tycon})=\taudef\\
+%	\taudef~((\Gamma; e)? :: (\Gamma; e_1)? :: \ldots :: (\Gamma; e_n)? :: []) \Downarrow_\fvalCtx \tden{\tau}{\iota}\\
+%			\trepof{\tau} \Downarrow_\fvalCtx \titype{\sigma}\\
+%	\Gamma \vdash_\fvalCtx \iota : \sigma
+%}{
+%	\atjsynX{e\cdot\FF{elim}[\tauidx](e_1; \ldots; e_n)}{\tau}{\iota}
+%}
+%\end{mathpar}
+%\caption{\small The active typing judgement. The normalization judgement for type-level terms ($\Downarrow$) and the representational consistency check will be in an appendix.}
+%\label{atj}
+%\end{figure}
+%\begin{figure}
+%\small
+%\begin{mathpar}
+%\inferrule[repof]{
+%	\tau \Downarrow_\fvalCtx \ttype{tycon}{\tauidx}\\
+%	\vdash_\fvalCtx \FF{rep}(\fvar{tycon}) = \taurep\\
+%	\taurep \tauidx \Downarrow_\fvalCtx \titype{\sigma}
+%}{
+%	\FF{repof}(\tau) \Downarrow_\fvalCtx \titype{\sigma}
+%}
+%
+%\inferrule[syn]{
+%    \tau \Downarrow_\fvalCtx (\Gamma; e)?\\
+%	\atjsynX{e}{\tau}{\iota}
+%}{
+%	\FF{syn}(\tau) \Downarrow_\fvalCtx \tden{\titerm{\iota}}{\tau}
+%}
+%
+%\inferrule[ana]{
+%	\tau_1 \Downarrow_\fvalCtx (\Gamma; e)?\\
+%	\tau_2 \Downarrow_\fvalCtx \ttype{tycon}{\tauidx}\\\\
+%	\atjanaX{e}{\ttype{tycon}{\tauidx}}{\iota}\\
+%}{
+%	\FF{ana}(\tau_1; \tau_2) \Downarrow_\fvalCtx \titerm{\iota}
+%}
+%\end{mathpar}
+%\caption{\small Evaluation semantics for the type-level language.
+Missing rules (including error propagation rules, which immediately cause failure of typechecking) are unsurprising and will be given in an appendix.}
+%\label{tleval}
+%\end{figure}
+%The expressive and metatheoretic power of the calculus arises from how the rules for the active typing judgement handle these  generalized forms.
+Rather than fixing the specification of a finite collection of operator constructors and tasking the \emph{compiler} with deciding a typing derivation on its basis, the specification instead delegates to a type-level function associated with a type constructor.
+In the core calculus, this is one of three functions, called \FF{iana}, \FF{isyn} and \FF{esyn} in the grammar.\footnote{This means that any particular type constructor only supports a single introduction and elimination desugaring.
+This is a minor inconvenience in some cases that is resolved in Ace by the use of methods.} We see how this is done with the rules for the active typing judgements, given in Figure \ref{atj}.
+%
+%The first two rules are standard in bidirectional type systems: variables synthesize types and if a term synthesizes a type, it can be analyzed against a type (cf.
+Listing \ref{context}).
+Variables translate to variables, and if we are simply converting from synthesis to analysis, translation is not affected.
+%
+%The rule $\fvar{att-lam}$ must take into account the fact that, because we support type-level computation, the type annotation on the argument may not be in normal form.
+Thus, we evaluate it to normal form.
+Note that the kinding rules (not shown) will guarantee that, because $\tau_1$ is of kind $\kTypeBlur$, its normal form, $\tau'_1$, is of the form $\ttype{tycon}{\tauidx}$ for some normal $\tauidx$.
+To generate an appropriate internal type annotation in the translation, we need to compute the representation type of $\tau_1'$.
+This involves calling the representation function associated with its type constructor (rule \fvar{repof} in Figure \ref{tleval}).
+The form $\titerm{\sigma}$ is a \emph{quoted internal type}.
+Note in the syntax that there is an unquote form, $\dup{\tau}$, which allows us to compose internal types compositionally without needing to expose an elimination form.
+Indeed, it is an interesting facet of our calculus that we never need to examine syntax trees directly to implement extensions.
+The evaluation semantics remove quotations, so the normal form of a quoted internal type contains an internal type with no quotations.
+%
+%Lambda functions are the only introductory form requiring special support in the calculus (because they need to  manipulate the context; see Discussion).
+The next three rules show how any other abstractions that we define (e.g.
+records, decimals, etc.) make use of a generalized introductory form.
+%
+%The rule  $\fvar{att-i-unasc}$ shows that unascribed introductory forms can only be analyzed against a type.
+Given such a type, the rule extracts a definition, named $\FF{iana}$, from the type constructor (cf.
+the \verb|ana_Dict| and \verb|trans_Dict| methods earlier).
+It calls this function with a pair containing the operator and type index and a list of \emph{reified arguments}, which have kind $\Q$ and introductory form $(\Gamma; e)?$.
+These are only constructed by the compiler (there would be no corresponding form in the concrete syntax).
+Their purpose is to allow the definition to programatically invoke synthesis and analysis as needed using the $\FF{ana}$ and $\FF{syn}$ operators.
+We can see the $\fvar{record}$ type constructor doing so on line 5 of Figure \ref{record-theory} to ensure that the field value provided as an argument has the same type as the corresponding label (accomplished by simultaneously folding over all three pieces of input data).
+The rule for performing analysis, $\fvar{ana}$, is in Figure \ref{tleval}.
+If it succeeds, analysis returns a \emph{translation}, which is a quoted internal term with kind $\kITerm$ and introductory form $\titerm{\iota}$.
+Like quoted internal types, there is an unquote form that is eliminated during evaluation.
+The operator definition uses this to construct a translation for the record.
+As before, empty records translate to units and records with a single field are unadorned.
+In this example, records with two or more fields translate to nested tuples.
+%
+%Because we have a representation type associated with the type, we can check that the translation is \emph{representationally consistent} (the final premise).
+As we will discuss, representational consistency combined with type safety of the internal language implies type safety overall (it arises as a strengthening of the inductive hypothesis needed to prove type safety, and is closely related to work on \emph{type-preserving compilation} in the TIL compiler for Standard ML, \cite{tarditi+:til-OLD}).
+%
+%The next rule, $\fvar{att-i-asc-ty}$ states that introductory forms ascribed with a type are analyzed against that type.
+%
+%Introductory forms ascribed with a type constructor can, according to the final introductory rule, $\fvar{att-i-asc-tycon}$, synthesize a type via the definition $\FF{isyn}$.
+It is passed the operator index (there is no type index, since we only have a type constructor) and a list of arguments, as before.
+In this case, it must return not just a translation, but also a type.
+We use the form $\tden{\tau_2}{\tau_1}$ for such a pairing, which has kind $\FF{TT}$.
+The type and translation can be extracted from it using the appropriately named elimination forms (indeed, as given, it is merely a pair, but we give it special syntax for clarity -- it looks like the conclusion of the typing judgement -- and for other reasons that will become clear in future work).
+To synthesize a type from a list of labels and arguments, we must be able to synthesize types from the arguments.
+The $\FF{syn}(\tau)$ operator permits this, per rule $\fvar{syn}$, seen being used in Listing \ref{record-theory}.
+We check representational consistency of the result, as before.
+%
+%Finally, we show the rule for elimination forms.
+It operates by first synthesizing a type for the ``primary'' subterm, then extracting the $\FF{esyn}$ definition from its type constructor.
+As before, it is called with the arguments and representational consistency is checked.
+Note that the primary subterm is itself the first argument (though we have already synthesized a type for it, it is more uniform and clear to allow the operator  to do so again, which is done in our example by the $\tvar{arity1}$ helper function).
+%
+\subsubsection{Representational Consistency}
+The representational consistency lemmas says that active typechecking and translation of an external term, if it succeeds (producing a type, $\hat\sigma$ and a translation $\iota$), always produces a well-typed translation.
+More specifically, the translation's type is the representation type of $\hat\sigma$.
+
+\begin{theorem}[Representational Consistency]
+Given well-kinded contexts and an external term that is both well-kinded and well-typed:
+\begin{enumerate}
+\item $\vdash \Phi$ 
+\item $\vdash_\Phi \Gamma$
+\item $\vdash_\Phi e$
+\item Either:
+\begin{enumerate}
+\item[a.] $\Gamma \vdash_\Phi e \Leftarrow \hat{\sigma} \leadsto \iota$
+\item[b.] $\Gamma \vdash_\Phi e \Rightarrow \hat{\sigma} \leadsto \iota$
+\end{enumerate}
+\end{enumerate}
+there exists an internal typing context, $\Omega$, and an internal type, $\tau$, such that  $\vdash_\Phi \Gamma \leadsto \Omega$ and $\FF{rep}(\hat\sigma) \Downarrow_\Phi \titype{\tau}$ and $\Omega \vdash \iota : \tau$.
+\end{theorem}
+\begin{proof}
+The proof proceeds by induction on the typing derivation (4a or 4b).
+Because the representational consistency is explicitly checked in the rules for intro and elim forms, these cases follow directly.
+The remaining cases are standard and follow inductively, given standard lemmas about valid contexts and a lemma about type safety of the static language.
+This is a fixed, standard functional language with only a few simple additions (TODO detail this).
+\end{proof}
+
+\subsubsection{Type Safety}
+Representational safety implies that well-typed external terms produce well-typed internal terms.
+If the internal language is type safe, then we know that well-typed external terms do not ``go wrong''.
+The internal language is simply PCF with sums and products, and a base type of integers, so this is a completely standard result \cite{pfpl}.
+
+%\subsubsection{Decidability}
+%The metatheoretic properties of interest are: type safety and termination of the type-level language (guaranteed by our kind system, though the details will need to be provided in an appendix; termination is non-trivial), type safety and decidability of the internal language (it i.s a standard variant of PCF, so this is trivial) and representational consistency.
 Type safety for the language as a whole comes as a corollary of these lemmas.
 We plan to provide detailed proofs of these, but for now, they should be treated as conjectures and our formulation above as expository.
-We believe that despite this, the formalization is surprisingly elegant and concise, given its expressive power.
+%We believe that despite this, the formalization is surprisingly elegant and concise, given its expressive power.
 It is a useful exercise to implement natural numbers, ala G\"odel's \textbf{T}, using this calculus, assuming that the type-level and internal languages include integers and one can lift them from the former into the latter.
@@
 
+TODO: mention this http://blog.codeclimate.com/blog/2014/05/06/gradual-type-checking-for-ruby/
+
 \section{Discussion}\label{discussion}

### 2014-05-10 ace-oopsla14/datatypes_t.py

@@ -1,15 +1,57 @@
-tree = lambda name, a: fp.data(name, 
-  lambda tree: {
-    'Empty': fp.unit,
-    'Leaf': a,
-    'Node': fp.tuple(tree, tree)
-  })
+import examples.fp
+__ace_base__ = fp
 
-dyntree = tree(dyn)
+@datatype
+def tree(a): [
+  Empty,
+  Node : (tree[a], tree[a]),
+  Leaf : a
+]
+
+def pair_trees(x, y):
+  {x, y : tree[a]]}
+  with (x, y).cases:
+    if Empty, Empty: 
+      Empty [:m,:p]
+    elif Node(x, y), Node(y): 
+      Node((x, y))
+    elif Leaf(x), Leaf(y): 
+      Leaf((x, y))
+    else:
+      raise
+
+
+
+
+@fp.datatype
+def x(a):
+  {'B': fp.unit,
+   'R': y}
+
+@fp.datatype
+def y(a):
+  {'B': fp.unit,
+   'R': x}
+
+@cl.struct
+def hello(): {
+  a : A,
+  b : B
+}
+
+@cl.struct
+def hello(): [
+  a : A,
+  b : B
+]
+
+
+   'Node': tuple(tree(a), tree(a))}
+@fp.datatype()
 
 @py
 def depth_gt_2(x):
-  {x : dyntree}
+  {x : tree[dyn]}
   return x.case({
     DT.Node(DT.Node(_), _): True,
     DT.Node(_, DT.Node(_)): True,

### 2014-05-10 ace-oopsla14/ace-oopsla14.tex
Fixed some things

@@
       return ctx.locals[x]
-    elif x in ctx.fn.static_env: # (Sec. 5)
+    elif x in ctx.fn.static_env:
       return self.syn_lifted(x)
@@
       return target.direct_translation(ctx, e)
-    else: # (Sec. 5)
+    else:
       return self.trans_lifted(ctx, target, e)

### 2014-06-04 atlang-gpce14/atlang-gpce14.tex
atlang-gpce14 submission

%\documentclass[12pt]{article}
\documentclass[9pt,preprint]{sigplanconf}
\newcommand{\sigmat}[1]{\sigma_{\text{#1}}}

\newcommand{\F}[1]{{\sf #1}~}
\newcommand{\FF}[1]{{\sf #1}}
\newcommand{\Q}{\FF{Arg}}
\usepackage{sfmath}
\usepackage{times}
%\setlength{\parskip}{0pt}
%\setlength{\parsep}{0pt}
\setlength{\headsep}{10pt}
%\setlength{\topskip}{0pt}
%\setlength{\topmargin}{0pt}
%\setlength{\topsep}{0pt}
%\setlength{\partopsep}{0pt}

\usepackage{subfigure}
\usepackage{adjustbox}
\usepackage{listings}
\usepackage[usenames,dvipsnames]{color}
\usepackage{mathpartir}
\usepackage{amsmath}
\usepackage{amssymb} 
%\usepackage{amsthm}
%\newtheorem{corollary}{Corollary}
\usepackage{ stmaryrd }
\usepackage{verbatimbox}
\input{../att-icfp14/macros-atlam}
\usepackage{alltt}
\renewcommand{\ttdefault}{txtt}
\usepackage{color}
\usepackage[usenames,dvipsnames,svgnames,table]{xcolor}
\definecolor{mauve}{rgb}{0.58,0,0.82}
\definecolor{light-gray}{gray}{0.5}
\usepackage{listings}
\usepackage{wasysym}
 \usepackage{longtable}
 \usepackage{booktabs}
 \usepackage{pdflscape}
 \usepackage{colortbl}%
 \newcommand{\myrowcolour}{\rowcolor[gray]{0.925}}
 \usepackage{wasysym}
\lstset{
  language=Python,
  showstringspaces=false,
  formfeed=\newpage,
  tabsize=4,
  commentstyle=\itshape\color{light-gray},
  basicstyle=\ttfamily\scriptsize,
  morekeywords={lambda, self, assert, cls, @classmethod, with},
  numbers=left,
  numberstyle=\scriptsize\color{light-gray}\textsf,
  numbersep=0.5em,
  xleftmargin=1.5em,
  stringstyle=\color{mauve},
  escapeinside={^}{^},
  aboveskip=1pt,
  belowskip=1pt
}
\lstdefinestyle{Bash}{
    language={}, 
    numbers=left,
    numberstyle=\scriptsize\color{light-gray}\textsf,
    moredelim=**[is][\color{blue}\bf\ttfamily]{`}{`}
}
\lstdefinestyle{OpenCL}{
  language=C++,
  morekeywords={kernel, __kernel, global, __global, size_t, get_global_id, sin, printf, int2}
}

\newcommand{\lstinlinep}[1]{\lstinline[language=Python,basicstyle=\ttfamily\small]{#1}}

\usepackage{float}
\floatstyle{ruled}
\newfloat{codelisting}{tp}{lop}
\floatname{codelisting}{Listing}
\setlength{\floatsep}{5pt}
\setlength{\textfloatsep}{6pt}
%\setlength{\intextsep}{0pt}
\setlength{\belowcaptionskip}{0pt}

\definecolor{very-light-gray}{gray}{0.9}

\usepackage{url}

\usepackage{todonotes}

\usepackage{placeins}

\usepackage{textpos}

\newtheorem{theorem}{Theorem}
\newtheorem{lemma}{Lemma}

\lefthyphenmin=62
\sloppy
%\makeatletter
%\renewcommand\section{\@startsection{section}{1}{\z@}%
%                       {-18\p@ \@plus -4\p@ \@minus -4\p@}%
%                       {8\p@ \@plus 4\p@ \@minus 4\p@}%
%                       {\normalfont\large\bfseries\boldmath
%                        \rightskip=\z@ \@plus 8em\pretolerance=10000 }}
%\renewcommand\subsection{\@startsection{subsection}{2}{\z@}%
%                       {-18\p@ \@plus -4\p@ \@minus -4\p@}%
%                       {8\p@ \@plus 4\p@ \@minus 4\p@}%
%                       {\normalfont\normalsize\bfseries\boldmath
%                        \rightskip=\z@ \@plus 8em\pretolerance=10000 }}
%\renewcommand\subsubsection{\@startsection{subsubsection}{3}{\z@}%
%                       {-5\p@ \@plus -4\p@ \@minus -4\p@}%
%                       {-0.5em \@plus -0.22em \@minus -0.1em}%
%                       {\normalfont\normalsize\bfseries\boldmath}}
%\renewcommand\paragraph{\@startsection{paragraph}{4}{\z@}%
%                       {-4\p@ \@plus -4\p@ \@minus -4\p@}%
%                       {-0.5em \@plus -0.22em \@minus -0.1em}%
%                       {\normalfont\normalsize\itshape}}
%\makeatother

    \renewcommand{\topfraction}{0.9}  % max fraction of floats at top
    \renewcommand{\bottomfraction}{0.8} % max fraction of floats at bottom
\newcommand{\rulename}[1]{{\textsc{\textbf{#1}}}}
\def \TirNameStyle #1{\small\rulename{#1}}
\begin{document}

\conferenceinfo{-}{-} 
\copyrightyear{-} 
\copyrightdata{[to be supplied]}

\title{Active Type Constructors}
\subtitle{Unambiguously Growing a Type System Inside a Python}
%\titlebanner{{\tt \textcolor{Red}{{\small Under Review -- distribute within CMU only.}}}}        % These are ignored unless
%\preprintfooter{Distribute within CMU only.}   % 'preprint' option specified.

%\authorinfo{~}{~}{~}
\authorinfo{Cyrus Omar \and Jonathan Aldrich}
          {          Carnegie Mellon University}
          {\{comar, aldrich\}@cs.cmu.edu}

\maketitle
\begin{abstract}
%Evidence suggests that programmers are reluctant to adopt new languages to
%gain access to new abstractions, even when they agree that these abstractions
%could be valuable. This suggests a need for languages that satisfy two key
%growth criteria: compatibility with existing programming ecosystems; and
%internal extensibility, so that new abstractions do not require new
%languages.
%
%We introduce @\texttt{lang}, a language compatible with the popular Python ecosystem.
%While Python, like other similar languages, is satisfactory for simple
%scripting, @\texttt{lang} is designed for more complex situations where static
%typechecking is beneficial. Unlike most statically-typed languages, @\texttt{lang}'s type
%system can be extended from within, by associating compile-time functions with
%type definitions. The compiler selectively invokes these according to a
%type-directed protocol, avoiding interference problems. Despite these design
%constraints, active types in @\texttt{lang} can express the static and dynamic semantics
%of a range of functional, object-oriented, parallel and domain-specific
%abstractions, all within libraries.
Programmers justifiably prefer abstractions available as libraries for an existing language over those requiring a switch to a new language, but library providers cannot extend these language's type systems directly, so powerful statically typed abstractions often remain obscure. This suggests a need for a language that is \emph{backwards compatible} with the libraries, tools and infrastructure of a widely adopted language but that has an  \emph{internally extensible type system}, organized around a mechanism that guarantees that any combination of extensions can be imported  without  ambiguity.

Our aim in this paper is thus to 1) introduce such a mechanism and 2) show how it can be implemented in a backwards compatible manner within an existing widely used language. We do so with two artifacts: @\texttt{lang}, an  extensible statically typed language implemented as a library inside Python, and @$\lambda$, a minimal calculus. Both @\texttt{lang} and @$\lambda$ have a fixed syntax but  determine the semantics of each term constructor by delegating control, according to a novel bidirectional typechecking and translation protocol, to methods associated with a relevant \emph{active type constructor}, defined in a library. The type constructor delegated control over a term is stable even when additional extensions are imported, so we avoid ambiguities by construction. % By flexibly  repurposing a fixed syntax in this type-directed manner, we avoid ambiguities by construction. %In @\texttt{lang}, types and functions are defined using Python's syntax, distributed using Python's package system and can call into Python libraries, so @\texttt{lang} is uniquely backwards compatible. We show that active type constructors as implemented in @\texttt{lang} permit practical library implementations of a variety of type system fragments that would normally need to be built into a language. %We also show how to construct safe, natural and extensible foreign function interfaces using a novel form of ``staged'' type propagation from Python into @\texttt{lang} and then through to OpenCL. % can be used to enable type system extensions and demonstrates how type safety can be maintained by lifting a technique from the typed compilation literature into the language.%To show that the method scales, we give a full implementation of the OpenCL programming language as a library. %We briefly examine the type theoretic foundations of this approach and give modular criteria that guarantee that an extension is safe and modular. %@\texttt{lang} can be used from the shell or interactively from within Python\todo{take this sentence out?}.
%Researchers developing languages and abstractions for high-performance computing must consider a number of design criteria, including performance, verifiability, portability and ease-of-use. Despite the deficiencies of legacy tools and the availability of seemingly superior options, end-users have been reluctant to adopt new language-based abstractions. We argue that this can be largely attributed to a failure to consider three additional criteria: continuity, extensibility\- and interoperability. This paper introduces @\texttt{lang}, a language that aims to satisfy this more comprehensive set of design criteria. To do so, @\texttt{lang} introduces several novel compile-time mechanisms, makes principled design choices, and builds upon existing standards in HPC, particularly Python and OpenCL. OpenCL support, rather than being built into the language, is implemented atop an extensibility mechanism that also admits abstractions drawn from other seemingly disparate paradigms. The core innovation underlying this and other features of @\texttt{lang} is a novel reification of types as first-class objects at compile-time, representing a refinement to the concept of active libraries that we call \emph{active types}. We validate our overall design by considering a case study of a simulation framework enabling the modular specification and efficient execution of ensembles of neural simulations across a cluster of GPUs.
\end{abstract}

%\category{D.3.2}{Programming Languages}{Language Classifications}[Extensible Languages]
%\category{D.3.4}{Programming Languages}{Processors}[Compilers]
%\category{F.3.1}{Logics \& Meanings of Programs}{Specifying and Verifying and Reasoning about Programs}[Specification Techniques]
%\keywords
%type-level computation, typed compilation
\section{Introduction}\label{intro}
Asking programmers to import a library is simpler than asking them to adopt a new programming language. Indeed, recent empirical studies underscore the difficulties of driving new languages into adoption, finding that extrinsic factors like compatibility with large existing code bases, library availability, familiarity and tool support are at least as important as intrinsic features of the language \cite{Meyerovich:2013:EAP:2509136.2509515,chen05}.

Unfortunately, researchers and domain experts designing new abstractions can  find it difficult to soundly and naturally implement them as libraries, particularly when these require strengthening the language's static type system. In these situations, abstraction providers generally develop a new language dialect, increasingly with the aid of tools like language workbenches \cite{erdweg2013state}. Unfortunately, this \emph{language-oriented approach} \cite{journals/stp/Ward94} is problematic at language boundaries: working with components written in a foreign language requires programming against (and thus revealing) its internal implementation. This is brittle, unnatural and unsafe and can cause performance issues due to additional necessary checks and conversions at language boundaries. %To avoid this, it is necessary to combine the different languages and their associated tools, but this often requires significant effort and expertise (we will discuss this further in Sec. \ref{related-work}).

These issues can prevent even dialects introducing only a few new constructs from being adopted. A study comparing a Java dialect, Habanero-Java (HJ), with a comparable library, \lstinlinep{java.util.concurrent}, for example, found that the language-based abstractions in HJ were easier to use and provided useful static guarantees \cite{cave2010comparing}. Nevertheless, it concluded that the library-based abstractions remained more practical outside the classroom because using HJ, as a distinct dialect of Java with its own type system, would be problematic in settings where some developers had not adopted it, but needed to interface with code that had. It would also be difficult to use it in combination with other abstractions also implemented as dialects of Java. Moreover, its tool support is more limited. 
This speaks to a widespread problem: programmers and development teams often cannot use  the typed abstractions they prefer because they are only available bundled with specialized languages they cannot  adopt \cite{Meyerovich:2012:SDR:2414721.2414724,Meyerovich:2013:EAP:2509136.2509515}. %These languages become, in essence, merely vehicles for communicating ideas, rather than useful tools.

 %This issue was perhaps most succinctly expressed by a participant in a recent study by Basili et al. \cite{basili2008understanding} who stated ``I hate MPI, I hate C++. [But] if I had to choose again, I would probably choose the same.''
%Unfortunately, researchers and domain experts (collectively, \emph{providers}) who design potentially useful new abstractions often find it difficult to implement them in terms of the general-purpose abstraction mechanisms available in existing languages. 
%Using a foreign construct can be  unnatural and, if the languages have  different type systems, can also cause safety problems.

Internally extensible languages have promised to reduce the problems associated with distinct dialects by giving abstraction providers more direct control over a language's syntax and semantics from within libraries. %In   an internally extensible language, programmers  can \emph{granularly} import the primitive constructs best suited to each component of their application or library. Providers can thus more easily develop, deploy and evaluate new abstractions in the context of existing codebases, narrowing one of the gaps between research and practice \cite{basili2008understanding}. 
%Indeed, abstractions could compete on their individual merit, rather than  by the serious adoption barriers mentioned above.
Unfortunately, the mechanisms available today have several problems. First, they are themselves often available only within a distinct dialect of an existing language and so face the classic ``chicken-and-egg'' problem: a language like SugarJ \cite{erdweg2011sugarj} (which permits syntax extensions over a fixed semantics) must overcome the same extrinsic barriers as a language like HJ. Second, giving abstraction providers too much control over the language can permit the definition of ambiguous extensions, a problem that is detected ``too late'': when extensions are combined by abstraction clients, who are not necessarily  capable of understanding and fixing the ambiguity.% Too many constraints can leave it difficult to express real-world abstractions.%It is difficult to guarantee that extensions are sound in isolation and more so to guarantee that they can be {safely} combined. %This requires further research before it would be appropriate to widely rely on them (which we will discuss as we go on).

\paragraph{Terminology}
A type system is typically defined in fragments organized around (and often identified by) a \emph{type constructor} (e.g. a fragment defining binary product types, organized around the type constructor $\fvar{prod}$). The syntax of each fragment can be characterized by its contributions to the \emph{concrete syntax} (e.g. expression forms \lstinlinep{(e, e)}, \lstinlinep{e.l} and \lstinlinep{e.r} and type form \lstinlinep{T * T}) and the \emph{abstract syntax}, where terms are written uniformly by applying a \emph{term constructor} to an \emph{index} and zero or more \emph{arguments}, e.g. $\FF{pair}[\FF{triv}](e_1; e_2)$ or $\FF{prj}[\FF{left}](e)$. Term indices are evaluated, if necessary, in an earlier \emph{phase} than term arguments and trivial indices are often omitted for clarity of presentation. The indices of expression terms,  like these examples,  are \emph{static terms}, $\sigma$. \emph{Types}, $\tau$, can be seen as static terms constructed by static term constructor $\FF{ty}$, indexed by a type constructor and having another (possibly trivial) static term as an argument, called (by convention) the \emph{type index}, e.g. $\FF{ty}[\fvar{prod}](\FF{spair}(\tau_1; \tau_2))$, often  abbreviated $\fvar{prod}\texttt{[}\tau_1; \tau_2\texttt{]}$ (%or stylized $\tau_1 \times \tau_2$. %Static terms are also called \emph{type-level terms}.  
 often, the static terms are not made explicit but rather assumed to be in the ambient metatheory).%Types classify expressions according to the rules comprising the fragment's type system. A type safe language is one where evaluation of a well-typed expression cannot lead to undefined behavior or result in a value of a different type, greatly simplifying static reasoning about program behavior and eliminating the need for many run-time checks.

%In a monolithic language design, decisions about which fragments are to be combined to constitute a complete  language are made centrally, so the designer can ensure that the concrete syntax is unambiguous and that other metatheoretic concerns, like guarantees of type safety, are maintained. To decentralize control over these decisions, so that fragments can be defined as libraries and combined arbitrarily, these guarantees must be shown to hold for all possible combinations of fragments deemed individually valid. This can quickly lead to difficulties.

  Each fragment also defines rules that contribute to the \emph{static semantics} (specifying how types are assigned to expressions at ``compile-time'') and the \emph{dynamic semantics} (specifying the behavior of well-typed expressions at ``run-time'') \cite{pfpl}. The dynamic semantics of full-scale languages are generally implemented (or even directly specified, as in the Harper-Stone semantics for Standard ML \cite{Harper00atype-theoretic}) by elaboration to a simpler \emph{internal language}.% In some languages, static terms themselves may have their own static and dynamic semantics (the \emph{static statics} and the \emph{static dynamics}).

The most apparently difficult problems related to combining fragments to form a typed programming language arise when attempting to enable the decentralized introduction of new concrete forms or term constructors. Syntactic ambiguities are difficult to modularly preclude and many tools operate by exhaustive case analysis over the concrete or abstract syntax of the terms of the language they work with (e.g. syntax highlighters, editor modes, pretty-printers, style checkers, documentation generators). Indeed, enabling the decentralized introduction of new term constructors is the standard example used when describing the more general problem of enabling the extension of a sum type with new cases as well as new  functions over its cases  \cite{Reynolds75}, prompting Wadler to dub it the \emph{expression problem} \cite{wadler1998expression}.

\paragraph{Approach} Fewer tools operate by exhaustive case analysis over type constructors. In a conventionally formulated programming language, however, it is difficult to introduce a new type system fragment without adding new concrete syntax or term constructors to accompany the new type constructors. In this paper, we aim to show that this need not be the case. We introduce a method that allows library providers to introduce new type system fragments by defining new type constructors in libraries while leaving the concrete syntax, the term constructors and the internal language fixed and enriching the static language so that it has its own dynamic semantics (the \emph{static dynamics}). The key idea is that we determine the semantics of each term not based directly on the identity of its term constructor (the usual ``syntax-directed'' approach), but rather by delegating to a static function associated with a relevant type constructor. The term constructor's identity only serves to determine how the delegate is assigned. We call user-defined type constructors equipped with static functions and distributed in libraries \emph{active type constructors}.

As a simple example, consider the concrete term \lstinlinep{e.r} (where \lstinlinep{e} is an expression and \lstinlinep{r} is a static label) corresponding to the now generalized abstract term  $\FF{attr}[\text{``r''}](e)$. Our semantics delegate responsibility for this term to a static function associated with the type constructor of the type recursively synthesized by the ``target'', $e$. As a result, tuple projection, as described above, can be introduced without the need to consider that it might syntactically coincide with projection out of a functional record or an object with a field labeled \lstinlinep{r} (both examples we will discuss). Each of these abstractions can use the same form and term constructor. As we will detail, terms for which there is no clear ``target'' from which to determine a delegate (e.g. number or list literals) instead either use the type they are being analyzed against or, if not available, a ``default'' delegate determined by the type constructor governing the function in which they appear. Symmetric binary terms (e.g. addition) use a third protocol, which we will discuss. There is always exactly one active type constructor delegated  responsibility over a term and this delegate is stable with respect to additional imports, so semantic ambiguities cannot arise by construction. %This is a generalization of a previous mechanism that we have developed, which deals only with extending the elaboration, but not the typechecking, of the bodies of delimited literals \cite{omar-ecoop14}.

%The approach we describe here is general with respect to the fixed syntax of the language it is introduced within. To implement it, we must only 1) define the delegation protocol that determines how a type constructor is delegated responsibility over each form, basing it on a bidirectional type system that distinguishes locations where an expression must \emph{synthesize} a type (e.g. to the right of bindings) from those where it can be \emph{analyzed} against a known type (this has also been called ``local type inference'' \cite{Pierce:2000:LTI:345099.345100}); 2) expose a fixed static language, which is used to implement active type constructors, construct types by applying type constructors to static indices and  import libraries; and 3) expose a fixed internal target language to the static language (fragment providers  implement the dynamic semantics of their fragments by translation to this language).

\paragraph{Specific Aims} Our specific aims in this paper are to introduce active type constructors and demonstrate that they are expressive, practically realizable within existing languages and theoretically well-founded. Our target audience is typed language fragment providers (researchers or domain experts) and language designers interested in extensibility mechanisms. We do not aim to evaluate the claim that our approach will ultimately be of use to fragment clients (i.e. normal developers) in writing and reasoning about a variety of real-world programs (this first requires adoption by our target audience). We describe two artifacts in pursuit of these aims:
\begin{enumerate}
\item Because we are committed to a fixed syntax, but aim to show that our mechanism does not critically rely on any particular syntactic choices, it is pragmatic to use an existing syntax for which tooling already exists. In this paper, we choose Python's syntax. Moreover, we use the Python language itself as both the static language and the internal language. This makes it easy for us to implement our language, called @\texttt{lang}, itself as a library inside Python, using its quasiquotation and reflection facilities.\footnote{The name was chosen as an initialism for ``actively  typed language'' and because Python marks function decorators, which we use extensively for this purpose, using the @ symbol.} Python is also an interesting starting point  because it  can be thought of as beginning with a degenerate static type system providing just one trivially indexed type constructor, \texttt{dyn} \cite{pfpl}. Taken together, @\texttt{lang} is almost completely backwards compatible with the Python ecosystem with remarkably little effort: we inherit a large number of  existing tools for working with Python files, Python's substantial packaging infrastructure and access to all existing Python libraries. We introduce @\texttt{lang} with practical examples of fragments as libraries in Sec. \ref{usage}.
\item In Sec. \ref{theory}, we describe active type constructors in their essential form, developing a minimal calculus called @$\lambda$. The syntax provides only variables, variable binding constructs, type ascription and two generic term constructors. Perhaps surprisingly, we show how a conventional syntax like Python's can be recovered by a purely syntactic desugaring to these term constructors. The semantics for @$\lambda$ take the form of a \emph{bidirectionally typed elaboration semantics} where the available type constructors are tracked by a \emph{fragment context}. We use a simple lambda calculus as both our static and internal language to emphasize the fundamental character of the approach.
\end{enumerate}
%@\texttt{lang} can be used from the shell, producing source files that can be further compiled or executed by external means. @\texttt{lang} functions targeting a language with Python bindings (e.g. Python itself, as well as the others just described) can also be compiled and invoked interactively from Python. 
%@\texttt{lang} supports a form of staged \emph{ad hoc} polymorphism (based on singleton types) wherein the ``dynamic type'' of a Python value can determine a static type and trigger active typechecking and translation ``just-in-time'' when Python code calls a generic @\texttt{lang} function. We show how this can significantly streamline interactive workflows (e.g. in scientific computing) that use Python for orchestration and exploratory tasks (e.g. plotting) but call into a statically-typed  language via a foreign-function interface (FFI) for performance- and safety-critical portions. We provide a complete implementation of the OpenCL type system (a language for low-level data parallel programming on GPUs), show several higher-level extensions to it, and discuss bindings that bridge between the type systems of @\texttt{lang}, OpenCL and Python's standard numerics package, \lstinlinep{numpy}, all as a library (contribution 3, Sec. \ref{targets}).%We will discuss in detail as our primary realistic case study.

%The remainder of the paper is organized as follows: in Sec. \ref{usage}, we describe the basic structure and usage of @\texttt{lang} with an example that uses several interesting @\texttt{lang} type libraries. In Sec. \ref{att}, we detail how active typing and translation delegate control, at compile-time, to active types and bases, and discuss how this represents a form of extensibility that side-steps the expression problem. In Sec. \ref{theory}, we reduce this mechanism to  a core lambda calculus, $\lam@\texttt{lang}$,  clarify the relationship to   type-level computation and show how type safety can be maintained. In Sec. \ref{targets}, we discuss generic functions, FFIs and staging via a complete implementation of the OpenCL type system (and extensions to it) as a library.

In Sec.  \ref{related}, we describe how active type constructors relate to several threads of related work. We conclude in Sec. \ref{discussion} by summarizing the key features needed by a host language to practically support active type constructors, and delineating present limitations and promising directions for future work. %The appendix has details omitted for concision as well as several more examples, including functional datatypes, flexible complex numbers and a previously little-explored type system for tracking string invariants using regular expressions statically (which we show being used in Listing \ref{example}, amongst  other examples).

\section{Active Type Constructors in @\texttt{lang}}\label{usage}

%
Listing \ref{example} shows an example of a well-typed @\texttt{lang} program for which strong correctness guarantees have been statically maintained using several type constructors defined in libraries. As suggested by line \ref{hello-ct}, the top level of every @\texttt{lang} file can be seen as a \emph{compilation script} written directly in Python, the static language. @\texttt{lang} requires no modifications to the language (version 2.6+ or 3.0+) or features specific to the primary implementation, CPython (so @\texttt{lang} supports alternative implementations like Jython and PyPy). This choice pays off immediately on the first line: @\texttt{lang}'s import mechanism is Python's import mechanism, so Python's build tools (e.g. \lstinlinep{pip}) and package repostories (e.g. \lstinlinep{PyPI}) are directly available for distributing @\texttt{lang} libraries, including those defining type constructors. Here, \lstinlinep{atlib} is the ``standard library'' but does not benefit from special support in \lstinlinep{atlang} itself.% Nothing in \lstinlinep{atlib} benefits from special support from \lstinlinep{atlang}, the core language package. We will see \lstinlinep{atlang} being used directly below. %@\texttt{lang} can be installed with a single command: \emph{omitted for review}. %@\texttt{lang} is agnostic about the particular design choices that we made in our examples.

%The top-level statements in an @\texttt{lang} file, like the \lstinlinep{print} statement on line 10, are executed to control the compile-time behavior, rather than the run-time behavior, of the program. %That is, Python serves as the \emph{compile-time metalanguage} (and, as we will see shortly, the \emph{type-level language}) of @\texttt{lang}. %For readers familiar with C/C++, Python can be thought of as serving a role similar to (but more general than) its preprocessor and template system (as we will see).
\subsection{Active Typechecking}

\paragraph{Types}
\begin{codelisting}[t]
\begin{lstlisting}
^\label{imports-start}^from atlib import record, decimal, dyn, string, 
  string_in, proto, fn, printf^\label{imports-end}^

^\label{hello-ct}^print "Hello, compile-time world!"

^\label{decl-Details}^Details = record[
  'amount' : decimal[2], 
  'memo' : dyn]
^\label{decl-Account-start}^Account = record[
  'name' : string, 
  'account_num' : string_in[r'\d{2}-\d{8}']]^\label{decl-Account-end}^
^\label{decl-Transfer}^Transfer = proto[Details, Account]

^\label{log-transfer-start}^@fn[Transfer, ...]
def log_transfer(t):
  """Logs a transfer to the console."""
  printf("Transferring %s to %s.", 
    t.amount.to_string, t.name)^\label{log-transfer-end}^

^\label{main-start}^@fn[...]
def __main__():
  ^\label{account-start}^account = {name: "Annie Ace", 
    account_num: "00-00000001"} [:Account]^\label{account-end}^
  log_transfer(({amount: 5.50, memo: None}, account))
  log_transfer(({amount: 15.00, memo: "Rent payment"}, 
    account))^\label{main-end}^

print "Goodbye, compile-time world!"
\end{lstlisting}
\caption{[\texttt{listing\ref{example}.py}] An @\texttt{lang} compilation script.}
\label{example}
\end{codelisting}
Types are constructed programmatically in the static language by applying a type constructor to an index:  \lstinlinep{tycon[idx]}. This is in contrast to many statically-typed languages where types (e.g. datatypes, interfaces) cannot be constructed computationally, but are only declared ``literally''. %Put another way, 
%@\texttt{lang} supports \emph{static computation at the level of types} and Python is its type-level language. The static language subsumes th%The compilation script can assign shorter names to types for convenience (removing the need for  facilities like \lstinlinep{typedef} in C or \lstinlinep{type} in Haskell). 
In this example, we see several useful types being constructed:
\begin{enumerate}
\item On line \ref{decl-Details}, we construct a functional record type with two (immutable) fields, assigning it to the identifier \lstinlinep{Details}. The field \lstinlinep{amount} has type \lstinlinep{decimal[2]}, which classifies decimal numbers having two decimal places, and \lstinlinep{memo} has type \lstinlinep{dyn}, classifying dynamic Python values.  %Here, \lstinlinep{record} and \lstinlinep{decimal} are {indexed type constructors}. 
Syntactically, the index for \lstinlinep{record} is provided using borrowed Python  syntax for array slices (e.g. \lstinlinep{a[min:max]}) to approximate conventional notation for type annotations. The field names are simply static strings (and could be computed rather than written literally, e.g. \lstinline{'a' + 'mount'}, to support features like \emph{type providers} \cite{Syme:2013:TIF:2429376.2429378}). 
\item On line \ref{decl-Account-start}, we construct another record type. The field \lstinlinep{name} has type \lstinlinep{string} and the field \lstinlinep{account_num} has a type classifying strings guaranteed statically to be in the regular language specified  by the  regular expression pattern provided as the index (written here as a \emph{raw string literal} \cite{python}). This  fragment, based on recent work, statically tracks the language an immutable string is in, providing operations like concatenation, demonstrating the expressive power of our approach \cite{fulton-thesis}. %To our knowledge, this type system is not  expressible in other simply typed languages (we do not consider dependently-typed logics without a distinct phase separation).
\item 
%Prototypic inheritance  is a feature of some dynamically-typed languages (most notably, Self \cite{Ungar:Smith:oopsla:1987} and Javascript). 
On line \ref{decl-Transfer}, we construct a simple \emph{prototypic object type} \cite{Lie86}, \lstinlinep{Transfer}, classifying values consisting of a \emph{fore} of type \lstinlinep{Details} and a \emph{prototype} of type \lstinlinep{Account}. %\footnote{We encourage the convention of naming types that are semantic in camelcase and types and type constructors that are structural in nature, like \texttt{dyn} or \texttt{record}, in lowercase.} 
 If a field cannot be found in the fore, the type system will delegate (statically) to the type of the prototype. This makes it easy to share the values of the fields of a common prototype amongst many fores, here to allow us to describe multiple transfers to the same account.% differing only in the transaction  details.
\end{enumerate}

\paragraph{Incomplete Types} Incomplete types arise from the application of a type constructor to an index containing an ellipsis (a valid array slice form in Python). The static terms \lstinlinep{fn[Transfer, ...]} and \lstinlinep{fn[...]} seen on lines \ref{log-transfer-start} and \ref{main-start} are incomplete types. We will discuss how the elided portion of the index (e.g. the return type) is determined below.

\paragraph{Active Type Constructors} Type constructors are implemented in @\texttt{lang} libraries as Python classes inheriting from \lstinlinep{atlang.Type} (classes here serving as Python's form of open sums). We show portions of \lstinlinep{atlib.fn} in Fig. \ref{fn-top}. Types are instances of these classes. Incomplete types are instead instances of \lstinlinep{atlang.IncompleteType}. The compiler controls instantiation by overloading the subscript operator on the \lstinlinep{atlang.Type} class object (using Python's \emph{metaclass} mechanism \cite{python}).

\paragraph{Function Definitions}
\begin{codelisting}[t]
\begin{lstlisting}
class fn(atlang.Type):
  ^\label{fn_init}^def __init__(self, idx): 
    # called by atlang.Type via the [] operator 
    atlang.Type.__init__(self, 
      self._check_and_norm(idx))

  @classmethod
  ^\label{syn_idx_FunctionDef}^def syn_idx_FunctionDef(cls, ctx, inc_idx, node):
    arg_types = cls._check_and_norm_inc(inc_idx)
    if not hasattr(ctx, 'assn_ctx'): 
      ctx.assn_ctx = { }
    ctx.assn_ctx.update(zip(node.args, arg_types))
    for stmt in node.body: ctx.check(stmt)
    last_stmt = node.body[-1]
    if isinstance(last_stmt, ast.Expr): 
      rty = last_stmt.value.ty
    else: rty = unit
    return (arg_types, rty) # fully specified index

  ^\label{ana_FunctionDef}^def ana_FunctionDef(self, ctx, node):
    if not hasattr(ctx, 'assn_ctx'): 
      ctx.assn_ctx = { } # top-level functions
    ctx.assn_ctx[node.name] = self # recursion
    ctx.assn_ctx.update(zip(node.args, self.idx[0]))
    # all but last
    for stmt in node.body[0:-1]: ctx.check(stmt)
    last_stmt = node.body[-1]
    if isinstance(last_stmt, ast.Expr): 
      ctx.ana(last_stmt.value, self.idx[1])
    elif self.idx[1] == unit or self.idx[1] == dyn: 
      ctx.check(last_stmt)
    else: raise atlang.TypeError("...", last_stmt)
\end{lstlisting}
\caption{A portion of the type constructor \texttt{atlib.fn}.}
\label{fn-top}
\end{codelisting}
 To be typechecked by @\texttt{lang} and define run-time behavior, function definitions must have an \emph{ascription}, provided by \emph{decorating} them with either a type or, here, an incomplete type (see \cite{python} for a discussion of Python decorators; we again use Python's metaclasses and operator overloading to support the use of a class as a decorator). Ascribed function definitions do not share Python's semantics and indeed the underlying Python function is discarded immediately after its abstract syntax and static environment (its closure and the global dictionary of the Python module it was defined in) have been extracted by the compiler using Python's built-in reflection capabilities and \lstinlinep{inspect} and \lstinlinep{ast} packages \cite{python}. The \lstinlinep{ast} package defines Python's term constructors. For function literals (term constructor \lstinlinep{FunctionDef}), the @$\texttt{lang}$ compiler delegates to the type constructor of the ascription in one of two ways, depending on the ascription.

If the ascription is an incomplete type, the compiler invokes the class method \lstinlinep{syn_idx_FunctionDef}, seen on line \ref{syn_idx_FunctionDef}, with the \emph{compiler context} (an object that provides hooks into the compiler and a storage location for accumulating information during typechecking), the incomplete index and the syntax tree of the function definition. Here, this method 1) checks and normalizes the incomplete index (here, checking that the argument types are indeed types and turning empty and single arguments into a 0- or 1-tuples for uniformity; not shown); 2) adds the argument names and types to a field of the context that tracks the types of local assignments (initializing it first if it is a top-level function, as in our example); 3) asks the compiler, via the method \lstinlinep{ctx.check}, to typecheck each statement in the body (discussed below); 4) if the term constructor of the last statement is \lstinlinep{ast.Expr} (a top-level expression), then the type of this expression is the return type, otherwise it is \lstinlinep{unit}; and finally, 5) a fully specified index (and thus a type) is synthesized for the function as a whole, here equivalent to \lstinlinep{fn[Transfer, atlib.unit]}. Note that this choice of using the last expression as the implicit return value (and considering any statement-level term constructor other than \lstinlinep{Expr} to have a trivial value) is made in a library, not by the language itself.% A different function type constructor could be written to support \lstinlinep{return} statements (omitted for concision).% The choices is made by clients .

Were the ascription a (complete) type, the compiler would delegate to \lstinlinep{fn} by calling an instance method on it, \lstinlinep{ana_FunctionDef}, seen on line \ref{ana_FunctionDef}. This method proceeds similarly, but does not need to perform the first step and last steps (and does not need to be a class method) because the index was already determined when the type was constructed, so it can be accessed via \lstinlinep{self.idx}. If the final statement is an expression, is analyzed against the provided return type (see below). Otherwise, only \lstinlinep{unit} or \lstinlinep{dyn} are valid return types (again, a choice made in a library)%. If it is, the expression . % Note that in this example, recursive functions are only supported when the ascription is a type, to avoid complications related to synthesizing a type for a recursive function call.
%On lines \ref{log-transfer-start}-\ref{log-transfer-end}, we see the function \lstinlinep{log_transfer}. We follow Python conventions by starting with a documentation string  that informally specifies the behavior of the function (so standard documentation generators can be used). Before moving into the body, however, we  also give an \emph{argument signature} (line \ref{arg-sig}) stating that the type of \lstinlinep{t} is \lstinlinep{Transfer}, the prototypic object type defined above.

%This again repurposes Python's array slice syntax to approximate conventional notation for type annotations. % (and following conventions for documentation strings, which also describe ``types'' at the end). % The ``keyword'' \lstinlinep{type} was chosen because it is the name of a built-in function in Python (and thus would not be used in the way shown for any other purpose).

\paragraph{Statements}
\begin{codelisting}[t]
\begin{lstlisting}
#class fn(atlang.Type): (cont'd) 
  @classmethod
  def check_Assign_Name(cls, ctx, stmt):
    x, e = stmt.target.id, stmt.value # see ast
    if x in ctx.assn_ctx: ctx.ana(e, ctx.assn_ctx[x])
    else: ctx.assn_ctx[x] = ctx.syn(e)

  @classmethod
  ^\label{syn_Name}^def syn_Name(cls, ctx, e):
    try: return ctx.assn_ctx[e.id]
    except KeyError:
      try: return self._syn_lift(ctx.static_env[e.id])
      except KeyError: raise atlang.TypeError("...", e)

  ^\label{default_asc}^@classmethod
  def syn_default_asc_Str(cls, ctx, e): return dyn
\end{lstlisting}
\caption{Some forms in the body of a function delegate to the type constructor of the function they are defined within (via class methods during typechecking).}
\label{fn-stmts}
\end{codelisting}
The \lstinlinep{ctx.check} method is defined by the compiler. As is the pattern in this paper, the compiler delegates to an active type constructor based on the term constructor of the statement being checked. Most statement-level term constructors other than \lstinlinep{Expr} simply delegate to the type constructor of the function they are defined within by calling class methods of the form \lstinlinep{check_TermCon}, where \lstinlinep{TermCon} is a term constructor or combination of term constructors in a few cases where a finer distinction than what Python itself made was necessary, as defined by \lstinlinep{ast}.

For example, the class method \lstinlinep{check_Assign_Name}, seen in Listing \ref{fn-stmts}, is called for statements of the form \lstinlinep{name = expr}, as on line \ref{account-start} of Listing \ref{example}. In this examples, the assignables context (the \lstinlinep{assn_ctx} field of the compiler context) is consulted to determine whether the name being assigned to already has a type due to a prior assignment, in which case the expression is analyzed against that type using \lstinlinep{ctx.ana}. If not, the expression must synthesize a type, using \lstinlinep{ctx.syn}, and a binding is added. %This interpretation of the assignment form cannot cause the function it appears in to return so \lstinlinep{None} is returned (Python returns \lstinlinep{None} by default if nothing is explicitly returned) and the subsumption rule applies trivially, so we do not need to define \lstinlinep{ana_rty_Assign_Name} separately.

%Return statements, on the other, require separately defining return type synthesis and analysis logic. The class method \lstinlinep{syn_rty_Return} is called for return statements when a return type is not yet known. It must synthesize a return type based on the value being returned (defaulting to \lstinlinep{unit}). Subsequent return statements are processed by \lstinlinep{ana_rty_Return}, which checks the provided value against the known return type. If no value is provided, then the return type must be \lstinlinep{unit}. Note that this function does not need to return a value; if a \lstinlinep{atlang.TypeError} is not raised, then analysis succeeds.

\paragraph{Bidirectional Typechecking of Expressions} The methods \lstinlinep{ctx.ana} and \lstinlinep{ctx.syn} are also defined by the compiler and can be called by type constructors to request type analysis (when the expected type is known) and synthesis (when the type is an ``output''), respectively, for an expression. Once again, the compiler delegates to a type constructor by a protocol that depends on the term constructor, invoking methods of the form \lstinlinep{ana_TermCon} or \lstinlinep{syn_TermCon}. This represents a form of bidirectional type system (sometimes also called a \emph{local type inference system}) \cite{Pierce:2000:LTI:345099.345100}, and the standard subsumption principle applies: if analysis is requested and an \lstinlinep{ana_TermCon} method is not defined but a \lstinlinep{syn_TermCon} method is, then synthesis proceeds and then the synthesized type is checked for equality against the type provided for analysis. Type equality is defined by checking that the two type constructors are identical and that their indices are equal. Two types with different type constructors are never equal, to ensure that the burden of ensuring that typing respects type equality is local to a single type constructor.  No form of subtyping or implicit coercion is supported for the purposes of this paper (though we have designed a mechanism that similarly localizes reasoning to a single type constructor, we do not have room to precisely describe it here.)

\paragraph{Names} If the term constructor of an expression is \lstinlinep{ast.Name}, as when referring to a bound variable or an assignable location, then the type constructor governing the function the term appears in is delegated to via the class method \lstinlinep{syn_Name} (names, when used as expressions, must synthesize a type). We see the definition of this method for \lstinlinep{fn} starting on line \ref{syn_Name} in Listing \ref{fn-stmts}. A name synthesizes a type if it is either in the assignables context or if it is in the static environment. In the former case, the type that was synthesized when the assignment occurred (by \lstinlinep{syn_Assign_Name}) is returned. In the latter case, we must lift the static value to run-time. For the purposes of this paper, we support only other typed @\texttt{lang} functions, Python functions and classes (which are given type \lstinlinep{dyn}, and can only be called with values of type \lstinlinep{dyn}) and modules (which are given a \emph{singleton type} -- a type indexed by the module reference itself -- from which @\text{lang} functions and Python functions and classes can be accessed as attributes in the usual way, see below).%In the future, we plan to detail how the interface between Python and @\texttt{lang} can be enriched to support moving typed values in and out of untyped Python functions safely (by inserting checks, defined again by a type constructor, at the interface between typed and untyped code), but we do not have enough room to adequately detail this mechanism here (and it is tangential to our specific aims).
\paragraph{Literal Expressions and Ascriptions}
\begin{codelisting}[t]
\begin{lstlisting}
class record(atlang.Type):
  def __init__(self, idx): 
    # Sig is an unordered mapping from fields to types
    atlang.Type.__init__(self, Sig.from_slices(idx))

  ^\label{syn_idx_Dict}^@classmethod
  def syn_idx_Dict(self, ctx, partial_idx, e):
    if partial_idx != Ellipsis: 
      raise atlang.TypeError("...bad index...", e)
    idx = []
    for f, v in zip(e.keys, e.values):
      if isinstance(f, ast.Name): 
        idx.append(slice(f.id, ctx.syn(v)))
      else: raise atlang.TypeError("...", f)
    return idx

  ^\label{ana_Dict}^def ana_Dict(self, ctx, e):
    for f, v in zip(e.keys, e.values):
      if isinstance(f, ast.Name):
        if f.id in self.idx.fields: 
          ctx.ana(v, self.idx[f.id])
        else: raise atlang.TypeError("...", f)
      else: raise atlang.TypeError("...", f)
    if len(self.idx.fields) != len(e.keys): 
      raise atlang.TypeError("...missing field...", e)

  ^\label{syn_Attribute}^def syn_Attribute(self, ctx, e):
    if e.attr in self.idx: return self.idx[e.attr]
    else: raise atlang.TypeError("...", e)
\end{lstlisting}
%
%  def trans_type(self, target):
%    if target.supports(examples.fp.EagerNAryProdLR):
%      return target.EagerNAryProdLR()
%    else if isinstance(target, examples.py.Dyn):
%      return target.Dyn()
\caption{A portion of the \texttt{atlib.record} type constructor.}
\label{record}
\end{codelisting}
Python designates literal forms for dictionaries, tuples, lists, strings, numbers and lambda functions. In @\texttt{lang}, the type constructor delegated control over terms arising from any of these forms is a function of how the typechecker encounters the term.

If the term appears in an analytic position, the type constructor of the type it is being analyzed against is delegated control. We see this on line \ref{account-start} of Listing \ref{example}: the dictionary literal form appears in an analytic context, here because an explicit type ascription, \lstinlinep{[:Account]}, was provided. Note that the ascription again repurposes Python's array slice syntax (the start is, conveniently,  optional). The compiler invokes the \lstinlinep{ana_Dict} method on the type the literal is being analyzed against, which is defined by its type constructor, here \lstinlinep{atlib.record}, shown on line \ref{ana_Dict} of Listing \ref{record}. This method analyzes each field value in the literal against the type of the field, extracted from the type index (an unordered mapping from field names to their types, so that type equality is up to reordering). The various literal forms inside the outermost form thus do not need an ascription because they appear in positions where the type they will be analyzed against is provided by \lstinlinep{record}. For example, the value of the field \lstinlinep{account_num} delegates to \lstinlinep{string_in} via the \lstinlinep{ana_Str} method (not shown).

An ascription directly on a literal can also be an incomplete type. For example, we can write \lstinline[basicstyle=\ttfamily\small]{[x, y] [:matrix[...]]} or more concisely \lstinline[basicstyle=\ttfamily\small]{[x, y] [:matrix]} instead of \lstinline[basicstyle=\ttfamily\small]{[x, y] [:matrix[i32]]} when we know \lstinline[basicstyle=\ttfamily\small]{x} and \lstinline[basicstyle=\ttfamily\small]{y} synthesize the type \lstinline[basicstyle=\ttfamily\small]{i32}, because the type constructor can use this information to synthesize the appropriate index. The decorator \lstinline[basicstyle=\ttfamily\small]{@fn[Transfer, ...]}, discussed previously, can be seen as a partial type ascription on a statement-level function literal.\footnote{In Python 3.0+, syntax for annotating function definitions directly with type-like annotations was introduced, so the entire index can be synthesized.} Lambda expressions support the same ascriptions:
\vspace{3px}
\begin{lstlisting}[numbers=none, basicstyle=\ttfamily\scriptsize]
(lambda x, y: x + y) [:fn[(i32, i32), ...]]
\end{lstlisting}
%\vspace{3px}
\noindent
Records support partial type ascription as well, e.g.: 
\vspace{3px}
\begin{lstlisting}[numbers=none,basicstyle=\ttfamily\scriptsize]
{name: "Deepak Dynamic", age: "27"} [:record]
\end{lstlisting}
%\vspace{3px}
\noindent
The compiler delegates to the type constructor by a class method in these cases (just as we saw above with \lstinlinep{fn}). In this case, the class method \lstinlinep{syn_idx_Dict}, shown on line \ref{syn_idx_Dict} of Listing \ref{record}, would be called. Because no record index was provided, an index must be synthesized from the literal itself, and thus the values are not analyzed against a type, but must each synthesize a type.

This brings us to the situation where a literal appears in a synthetic position, as the two string literals above do. In such a situation, the compiler delegates responsibility to the type constructor of the function that the literal appears in, asking it to provide a ``default ascription'' by calling the class method \lstinlinep{syn_default_asc_Str}, shown on line \ref{default_asc} of Listing \ref{fn-stmts}. In this case, we simply return \lstinlinep{dyn}, so the type of the expression above has type \lstinline[basicstyle=\ttfamily\small]{record['name': dyn, 'age' : dyn]}. A different function type constructor might make more precise choices. Indeed, a benefit of using Python as our static language is that it is relatively straightforward to provide a type constructor that allows us to provide different defaults, and control choices like the return semantics, as block-scoped ''pragmas'' in the static language using Python's \lstinlinep{with} statement \cite{python} (details omitted), e.g. 
\begin{codelisting}[h]
\begin{lstlisting}[numbers=none]
with fn.default_asc(Num=i64, Str=string, Dict=record):
  @fn[...] # : fn[(), record["a": i64, "b": string]]
  def test(): {a: 1, b: "no ascriptions needed!"} 
\end{lstlisting}
\caption{Block-scoped settings can be seen by type constructors.}
\label{defaultasc}
\end{codelisting}
\vspace{-10px}
%{In Python 3, syntax for annotating functions (but unfortunately not lambdas) with argument and return type annotations directly  (purely as metadata) was introduced \cite{pep3107}:
%\begin{lstlisting}[numbers=none]
%@fn
%def f(x : T) -> T: # ...
%\end{lstlisting}
%\vspace{-3px}
%This can also be thought of as an incomplete type ascription, which we could write more explicitly \lstinlinep{@fn[...]},  because the literal itself contains the necessary information to synthesize the full index. We show the more universally available notation here because Python 2.\textit{x} is much more widely adopted as of this writing.

%\paragraph{Compound Expressions} 
%For expressions other than names and literals, the type synthesized by one or more subexpressions determines the type constructor delegated responsibility for synthesis and analysis.%The sub-expression that determines this is a function of the term constructor.

\paragraph{Targeted Expressions} Expression forms having exactly one sub-expression, like \lstinlinep{-e} (term constructor \lstinlinep{UnaryOp_USub}) or \lstinlinep{e.attr} (term constructor \lstinlinep{Attribute}), or where there may be multiple sub-expressions but the left-most one is the only one required, like \lstinlinep{e[slices]} (term constructor  \lstinlinep{Subscript}) or \lstinlinep{e(args)} (term constructor \lstinlinep{Call}), are called \emph{targeted expressions}. For these, the compiler first synthesizes a type for the left-most sub-expression, then calls either the \lstinlinep{ana_TermCon} or \lstinlinep{syn_TermCon} methods on that type. For example, we see type synthesis for the field projection operation on records defined starting on line \ref{syn_Attribute} of Listing \ref{record}.% We omit the implementation for concision, but \lstinlinep{fn} defines \lstinlinep{syn_Call} as one would expect, by simply checking each argument against the corresponding argument in the index.

\paragraph{Binary Expressions} \begin{codelisting}[t]
\begin{lstlisting}
class string_in(atlang.Type):
  def __init__(self, idx):
    atlang.Type.__init__(self, self._rlang_of_rx(idx))

  # treats string as string_in[".*"]
  ^\label{handles_Add_with}^handles_Add_with = set([string]) 
  ^\label{syn_BinOp_Add}^@classmethod
  def syn_BinOp_Add(cls, ctx, e): 
    rlang_left = self._rlang_from_ty(ctx.syn(e.left)) 
    rlang_right = self._rlang_from_ty(ctx.syn(e.right))
    return string_in[self._concatenate_langs(
      rlang_left, rlang_right)]
\end{lstlisting}
\caption{Binary operations in \texttt{atlib.string\_in}.}
\label{string_in}
\end{codelisting}
\begin{codelisting}[t]
\begin{lstlisting}
def check_tycon(tycon):
  for other_tycon in tycon.handles_Add_with: 
    if tycon in other_tycon.handles_Add_with:
      raise atlang.AmbiguousTyconError("...", 
        tycon, other_tycon) # (other binops analagous)
\end{lstlisting}
\caption{For each type constructor definition and binary operator, \texttt{atlang} runs a modular handle set check to preclude ambiguity.}
\label{check_tycon}
\end{codelisting}
The major remaining expression forms are the binary operations, e.g. \lstinlinep{e + e} or \lstinlinep{e < e}. These are notionally symmetric, so it would not be appropriate to simply give the left-most one precedence. Instead, we begin by attempting to synthesize a type for both subexpressions. If both synthesize a type with the same type constructor, or only one synthesizes a type, that type constructor is delegated responsibility via a class method, e.g. \lstinlinep{syn_BinOp_Add} on line \ref{syn_BinOp_Add} of Listing \ref{string_in}.

If both synthesize a type but with different type constructors (e.g. we want to concatenate a \lstinlinep{string} and a \lstinlinep{string_in[r]}), then we consult the \emph{handle sets} associated as a class attribute with each type constructor, e.g. \lstinlinep{handles_Add_with} on line \ref{handles_Add_with} of Listing \ref{string_in}. This is a set of other type constructors that the type constructor defining the handle set may potentially support binary operations with. When a type constructor is defined, the language checks that if \lstinlinep{tycon2} appears in \lstinlinep{tycon1}'s handler set, then \lstinlinep{tycon1} does \emph{not} appear in \lstinlinep{tycon2}'s handler set. This is a very simple modular analysis (rather than one that can only be performed at ``link-time''), shown in Listing \ref{check_tycon}, that ensures that the type constructor delegated control over each binary expression is deterministic and  unambiguous without arbitrarily preferring one subexpression position over another. This check is performed  by using a metaclass hook (technically, this can be disabled; we assume that clients are not importing potentially adversarial extensions in this work, though lifting this assumption raises quite interesting questions that we hope will be addressed by future work).

\subsection{Active Translation}
Once typechecking is complete, the compiler enters the translation phase. %The compiler creates an instance of a class inheriting from \lstinlinep{ace.Target} for use during this phase via the \lstinlinep{init_target} method (line \ref{pybase}.54). This object provides methods for code generation and supports features like fresh variable generation, adding imports and so on. It's interface is not constrained by @\texttt{lang} (we will see what @\texttt{lang} requires of a target below) and the mechanics of code generation are orthogonal to the  focus of this paper, so we will discuss it relatively abstractly. The simplest API would be string-based code generation. For the Python target we use here, we generate ASTs. The API is based directly on the \lstinlinep{ast} library, with a few additional conveniences just mentioned.
This phase follows the same delegation protocol as the typechecking phase. Each \lstinlinep{check_}/\lstinlinep{syn_}/\lstinlinep{ana_TermCon} method has a corresponding \lstinlinep{trans_TermCon} method. These are all now instance methods, because all indices have been fully determined.

Examples of translation methods for the \lstinlinep{fn} and \lstinlinep{record} type constructors are shown in Listing \ref{trans}. The output of translation on our example is discussed in the next subsection. Translation methods have access to the context and node, as during typechecking, and must return a translation, which for our purposes, is simply another Python syntax tree (in practice, we offer some additional conveniences beyond \lstinlinep{ast}, such as fresh variable generation and lifting of imports and statements inside expressions to appropriate positions, in the module \lstinlinep{astx} distributed with the language). Translation methods can assume that the term is well-typed and the typechecking phase saves the type that was delegated control, along with the type that was assigned, as attributes of each term processed by the compiler. Note that not all terms need to have been processed by the compiler if they were otherwise reinterpreted by a type constructor given control over a parent term (e.g. the field names in a record literal are never treated as expressions, while they would be for a dictionary literal). %This protocol can be seen on lines \ref{context}.39-\ref{context}.47. 
\begin{codelisting}[t]
\begin{lstlisting}
#class fn(atlang.Type): (cont'd)
  def trans_FunctionDef(self, ctx, node):
    x_body = [ctx.trans(stmt) for stmt in node.body]
    x_fun = astx.copy_with(node, body=x_body)
    if node.name == "__main__": 
      x_invoker = ast.parse(
        'if __name__ == "__main__": __main__()')
      return ast.Suite([x_fun, x_invoker])
    else: return x_fun

  def trans_Assign_Name(self, ctx, stmt):
    return astx.copy_with(stmt, 
      target=ctx.trans(stmt.target), 
      value=ctx.trans(stmt.value))

  def trans_Name(self, ctx, e):
    if e.id in ctx.assn_ctx: return astx.copy(e)
    else: return self._trans_lift(
      ctx.static_env[e.id])

#class record(atlang.Type): (cont'd)
  def trans_Dict(self, ctx, e): 
    if len(self.idx.fields) == 1: 
      return ctx.trans(e.values[0])
    ast_dict = dict(zip(e.keys, e.values))
    return target.Tuple(ctx.trans(target, ast_dict[f])
      for f, ty in self.idx)

  def trans_Attribute(self, ctx, e):
    if len(self.idx.fields) == 1: return ctx.trans(e)
    else: return ast.Subscript(ctx.trans(e.value), 
      ast.Index(ast.Num(self.idx.position_of(e.attr))))
\end{lstlisting}
\caption{Translation methods for the types defined above.}
\label{trans}
\end{codelisting}

%Because we are targeting Python directly, most of our \lstinlinep{trans} methods are direct translations (factored out into a helper function). The main methods of interest that are not entirely trivial are \lstinlinep{trans_FunctionDef_outer} on lines \ref{pybase}.33-\ref{pybase}-37 and the two translation methods in Listing \ref{record}, which implement the logic described in Sec. \ref{compilation}. 
%
%Each type constructor must also specify a \lstinlinep{trans_type} method. This is important when targeting a typed language (so that the translations of type annotations can be generated, for example). As we will see in the next two sections, this is also critical to ensuring type safety. The language \emph{checks} not only that translations having a given type are well-typed, but that they have the type specified by this method (requiring that the target provide a \lstinlinep{is_of_type} method). Here, because we are simply targeting Python directly, the \lstinlinep{trans_type} method on line \ref{record}.49-\ref{record}.50 simply generates \lstinlinep{target.dyn}. 
%
%When this phase is complete, each node processed by the context will have a translation, available via the \lstinlinep{trans} attribute. In particular, each typed function has a translation. Note that some nodes are never processed by the context because they were reinterpreted by the delegate (e.g. the field names in a record literal), so they do not have translations (as expected, given our discussion above).
%
%To support external compilation, the target must have an \lstinlinep{emit} method that takes the compilation script's file name and a reference to a string generator (an instance of \lstinlinep{ace.util.CG}) and emits source code. The string generator we provide can track indentation levels (to support Python code generation and make generation for other languages more readable, for the purposes of debugging). It allows non-local string generation via the concept of user-defined \emph{locations}.  Each file that needs to be generated is a location and there can also be locations within a file (e.g. the imports vs. the top-level code), specified by a target the first time it finds that a necessary location is not defined. A generated entity (e.g. an import, class definition or function definition) can only be added once at a location. We saw this in Listing \ref{example-out} for the decimal-to-string conversion. The API will be discussed further in the appendix.

\subsection{Standalone Compilation}\label{compilation} 
\begin{codelisting}[t]
\begin{lstlisting}[style=Bash]
$ `@ listing1.py`
Hello, compile-time world!
Goodbye, compile-time world!
[@] _atout_listing1.py successfully generated.
$ `python _atout_listing1.py`
Transferring 5.50 to Annie Ace.
Transferring 15.00 to Annie Ace.
\end{lstlisting}
\caption{Compiling \texttt{listing\ref{example}.py} using the \texttt{@} script.}
\label{external-compilation}
\end{codelisting}
\begin{codelisting}[t]
\begin{lstlisting}
import atlib.runtime as _at_i0_

def log_transfer(t):
  _at_i0_.print("Transferring %s to %s." % 
    (_at_i0_.dec_to_str(t[0][0], 2), t[1][0]))

def __main__():
  account = ("Annie Ace", "00-00000001")
  log_transfer((((5, 50), None), account))
  log_transfer((((15, 0), "Rent payment"), account))
if __name__ == "__main__": __main__()
\end{lstlisting}
\caption{[\texttt{\_atout\_listing\ref{example}.py}] The file generated in Listing \ref{external-compilation}.}
\label{example-out}
\end{codelisting}

Listing \ref{external-compilation} shows how to invoke the \lstinlinep{@} compiler at the shell to typecheck and translate then execute \lstinlinep{listing1.py}. %Note that the {\texttt{print}} statements at the top  of the  compilation script were evaluated during compilation only. %These two steps can be combined by running \lstinlinep{ace listing1.py} (the intermediate file is not generated unless explicitly requested in this case).
The @\texttt{lang} compiler is itself a Python library and \lstinlinep{@} is a simple Python script that invokes it in two steps:
\begin{enumerate}
\item It evaluates the compilation script to completion.
\item For any top-level bindings in the environment that are @\texttt{lang} functions, it initiates typehecking and translation as described above. If no type errors are discovered, the ordered set of translations are collected (obeying order dependencies) and emitted. If a type error is discovered, an error is displayed.
\end{enumerate}

In our example, there are no type errors, so the file shown in Listing \ref{example-out}  is generated. This file is meant only to be executed, not edited or imported directly. The invariants necessary to ensure that execution does not ``go wrong'', assuming the extensions were implemented correctly, were checked statically and entities having no bearing on execution, like field names and types themselves, were erased. The dynamic semantics of the type constructors used in the program were implemented by translation to Python:% In particular, note: 
\begin{enumerate}
\item \lstinlinep{fn} recognized the function name \lstinlinep{__main__} as special, inserting the standard Python code to invoke it if the file is run directly.
\item Records translated to tuples (the field names were not needed).
\item Decimals translated to pairs of integers. String conversion is defined in a ``runtime'' package with a ``fresh'' name.
\item Terms of type \lstinlinep{string_in[r"..."]} translated to strings. Checks could here be performed  statically.% Note that there are situations (e.g. if a string is read in from the console) that would necessitate an initial run-time check, but no further checks in downstream functions.
\item Prototypic objects translated to pairs consisting of the fore and the prototype. Dispatch to the appropriate record based on the field name was static (line 5).
\end{enumerate}

\paragraph{Type Errors} Listing \ref{oops} shows an example of code containing several type errors. If analagous code were written in Python itself, these could only be found if the code was executed on the first day of the month (and not all of the issues would  immediately be caught as run-time exceptions). In @\texttt{lang}, these can be found without executing the code (i.e. true static typechecking).

\begin{codelisting}[t]
\begin{lstlisting}
from listing1 import fn, dyn, Account, Details, 
  log_transfer
import datetime
@fn[dyn, dyn]
def pay_oopsie(memo):
  if datetime.date.today().day == 1: # @lang to Python
^\label{error-start}^    account = {nome: "Oopsie Daisy", 
      account_num: "0-00000000"} [:Account] # (format)
    details = {amount: None, memo: memo} [:Details]
    details.amount = 10 # (immutable)
    log_transfer((account, details)) # (backwards) ^\label{error-end}^
print "Today is day " + str(datetime.date.today())
pay_oopsie("Hope this works..") # Python to @lang
print "All done."
\end{lstlisting}
\caption{[\texttt{listing\ref{oops}.py}] Lines \ref{error-start}-\ref{error-end} each have a type error.}
\label{oops}
\end{codelisting}
\begin{codelisting}[t]
\begin{lstlisting}[style=Bash]
$ `python listing^\color{blue}\ref{oops}^.py`
Today is day 2
Traceback (most recent call last):
  File "listing^\ref{oops}^", line 9, in <module>
atlang.TypeError: 
    File "listing^\ref{oops}^.py", line ^\ref{error-start}^, col 12, in <module>: 
      [record] Invalid field name: nome
\end{lstlisting}
\caption{Execution never proceeds into a function with a type error.}
\label{oops-exec}
\end{codelisting}

\subsection{Interactive Invocation} @\texttt{lang} functions over values of type \lstinlinep{dyn} can be invoked directly from Python. Typechecking and translation occurs upon first invocation (subsequent calls are cached; we assume that the static environment is immutable). We see this occurring when we execute the code in Listing \ref{oops} using the Python interpreter in Listing \ref{oops-exec}.

In future work, we plan to detail how to insert dynamic checks and wrapper objects, defined by active type constructors in a manner similar to how static checks are defined here, so that values that are not of type \lstinlinep{dyn} can be passed in and out of untyped Python code. Because these can be seen as implicit coercions to and from \lstinlinep{dyn}, and we do not aim to introduce this feature in this paper, we omit discussion of this topic. Explicit coercions can be implemented using the mechanisms described above. For example, \lstinlinep{string_in} provides an introductory form that checks a provided \lstinlinep{string}  or \lstinlinep{dyn} value dynamically, raising an exception in the case of failure: \lstinlinep{[raw_input()] : string_in[r"\\d+"]}.

%.; for example, the unexpected use of \lstinlinep{None}\footnote{The notation \texttt{+T} constructs a \texttt{None}-able option type; see appendix.}). Static type checking allows us to find these errors during compilation. Listing \ref{oops-compilation} shows the result of attempting to compile this code. The compilation script completes (so that functions can refer to each other mutually recursively), then the typed functions in the top-level environment are typechecked. The typechecker raises an exception statically at the first error, and \lstinlinep{acec} prints it to the console as shown.

%When encountering a compound term (e.g. \lstinlinep{t.amount} in Listing \ref{example}), the compiler defers control over type  and translation to the type of a designated subexpression (e.g. \lstinlinep{t}) according to a fixed fixed \emph{dispatch protocol}. Below are examples of the choices made in @\texttt{lang}.\todo{full in appendix?}
%\begin{itemize}
%\item Responsibility over {\bf attribute access} (\texttt{e.attr}), {\bf subscripting} (\texttt{e[e1]}) and \textbf{calls} (\lstinlinep{e(e1, ..., en)}) and {\bf unary operations} (e.g. \lstinlinep{-e}) is handed to the type recursively assigned to \texttt{e}.
%\item Responsibility over {\bf binary operations} (e.g. \lstinlinep{e1 + e2}) is first handed to the type assigned to the left operand. If it indicates that it cannot handle the operation, the type assigned to the right operand is handed responsibility. {Note that this operates like the corresponding rule in Python's \emph{dynamic} operator overloading mechanism; see Sec. \ref{related} for a discussion.}
%\item Responsibility over \textbf{constructor calls} (\lstinlinep{[t](e1, ..., en)}), where \lstinlinep{t} is a \emph{compile-time Python expression} evaluating to an active type, is handed to that type. If \lstinlinep{t} evaluates to a \emph{family} of types, like \lstinlinep{clx.Cplx}, the active type is first generated via a class method, as discussed below.
%%\item Responsibility over {\bf simple assignment statements}, is handed to the type of the variable on the left (which, as we will see below, is determined by the active base). If this type does not provide special assignment semantics, the base must handle it.%Destructuring assignment is also supported by a somewhat more complex protocol.\todo{clarify}
%\end{itemize}
%%
%%\todo{To be revised from here on down}The core of @\texttt{lang} consists of about 1500 lines of Python code implementing its primary concepts: generic functions, concrete functions, active types, active bases and active targets.  The latter three comprise @\texttt{lang}'s extension mechanism. Extensions provide semantics to, and govern the compilation of, @\texttt{lang} functions, rather than logic in @\texttt{lang}'s core. %Indeed, the name ``@\texttt{lang}'' might itself be an acronym: it is an ``active compilation environment''.

% As a result, execution is essentially as fast as it could be given the target language chosen (Python). We will target an even faster language, \lstinlinep{OpenCL}, in Sec. \ref{targets}.

%Because \lstinlinep{t} will be statically guaranteed to have type \lstinlinep{Transfer},

\section{@$\lambda$: Active Type Constructors, Minimally}\label{theory}
\newcommand{\cev}[1]{\reflectbox{\ensuremath{\vec{\reflectbox{\ensuremath{#1}}}}}}
\newcommand{\tri}{i}
\newcommand{\trt}{\cev\tau}
\renewcommand{\tnil}[1]{[]}
\newcommand{\ltxt}[1]{\ell_{\text{#1}}}
\newcommand{\abslbl}[1]{\FF{lbl}[{\texttt{#1}}]}
\begin{figure*}[t!]
\small
$
\begin{array}{ll}
\F{import} \Phi_\text{nat}, \Phi_\text{lprod} & \FF{import}[\Phi_\text{nat}, \Phi_\text{lprod}](\\
\F{static~let} \tvar{nat} = \fvar{nat}\texttt{\textbf{[}}\FF{nil}\texttt{\textbf{]}} & \FF{slet}[\FF{ty}[\fvar{nat}](\FF{nil})](\tvar{nat}.\\
\F{let} zero = \texttt{0} : \tvar{nat} & \FF{let}(\FF{asc}[\tvar{nat}](\FF{lit}[\FF{lbl}[\texttt{0}]](\cdot)); zero.\\
\F{let} one = zero.{\texttt{s}} & \FF{let}(\FF{targ}[\abslbl{s}](zero; \cdot); one.\\
\F{let} plus = (\lambda x{:}\tvar{nat}. \lambda y{:}\tvar{nat}. &  \FF{let}(\FF{asc}[\FF{incty}[\fvar{fn}](\tvar{nat})](\lambda(x.\FF{asc}[\FF{incty}[\fvar{fn}](\tvar{nat})](\lambda(y.\\
\quad x\cdot{\texttt{rec}}(y; \lambda p.\lambda r. r\cdot{\texttt{s}})) & \quad \FF{targ}[\abslbl{rec}](x; \lambda(p.\lambda(r.\FF{targ}[\abslbl{s}](r; \cdot)))))))); plus.\\
\F{let} two = plus~one~one & \FF{let}(\FF{targ}[\FF{nil}](\FF{targ}[\FF{nil}](plus; one); one); two.\\
\{{\texttt{one}} {=} one; {\texttt{two}} {=} two\} : \fvar{lprod}\texttt{[}\cdots\texttt{]}~~~~~~& \FF{asc}[\FF{incty}[\fvar{lprod}](\FF{nil})](\FF{lit}[\FF{cons}(\abslbl{one}; \FF{cons}(\abslbl{two}; \FF{nil}))](one; two)))))))\\
% & \quad \\
%\quad\quad \\
%\quad\quad\quad & \quad\quad\quad \\
%\{{\texttt{one}} {=} one; {\texttt{two}} {=} plus(one)(one)\} : \fvar{lprod} ~~~~& \FF{asc}[\FF{incty}[\fvar{lprod}](\FF{nil})](\\&
%\quad 
%one; \\&
%\quad \FF{targ}[\FF{nil}](\FF{targ}[\FF{nil}](plus; one); one))))))))
\end{array}
$~~
%$\begin{array}{rcl}
%\fvar{tycon}\texttt{[}\sigma\texttt{]} & := & \FF{ty}[\fvar{tycon}](\sigma)\\
%\fvar{tycon}\texttt{[}\cdots\sigma\cdots\texttt{]} & := & \FF{incty}[\fvar{tycon}](\sigma)\end{array}$
%~~~~~~~~$
%\begin{array}{rcl}
%\fvar{tycon} & := & \FF{incty}[\fvar{tycon}](\FF{nil})\\
%\fvar{tycon}\texttt{[}\cdots\texttt{]} & := & \FF{incty}[\fvar{tycon}](\FF{nil})\\
%\end{array}$\\
%~\\
%\hspace{-50px}$
%\tiny
%\begin{array}{rcl}
%\lambda x{:}\sigma.e & := & \FF{asc}[\FF{incty}[\fvar{fn}](\sigma)](\lambda(x.e))\\
%(e_1, ..., e_n) & := & \FF{lit}[\FF{nil}](e_1; \ldots; e_n)\\
%\{{\texttt{lbl}}_1=e_1, \ldots, {\texttt{lbl}}_n=e_n\} & := & \FF{lit}[\FF{cons}(\FF{lbl}[{\texttt{lbl}}_1]; \ldots\FF{cons}(\FF{lbl}[{\texttt{lbl}}_n]; \FF{nil}))](e_1; \ldots; e_n)\\
%\text{(any numeric label)}~~~~{\texttt{n}} & := & \FF{lit}[\FF{lbl}[\texttt{n}]]()\\
%e(e_1) & := & \FF{targ}[\FF{nil}](e; e_1)\\
%e\cdot{\texttt{lbl}} & := & \FF{targ}[\FF{lbl}[{\texttt{lbl}}]](e; \cdot)\\
%e\cdot{\texttt{lbl}}(e_1, ..., e_n) & := & \FF{targ}[\FF{lbl}[{\texttt{lbl}}]](e; e_1; \ldots; e_n)
%%%(e_1, ..., e_n) & := & intro[()](e_1; ...; e_n)
%%%n &  := & intro[n]()
%%%s & := & intro[s]()
%%%
%%%e.l & := & elim[l](e)
%%%e[e_1; ...; e_n] & := & elim[()](e_1, ..., e_n)
%%%e.l(e_1; ...; e_n) & := & elim[l]
%\end{array}
%$
%
\caption{A program written using conventional concrete syntax, left, syntactically desugared to the abstract syntax on the right.}
\label{atlam-example}
\end{figure*}
\begin{figure}[t!]
\newcommand{\eqval}[1]{#1}
\newcommand{\noneqval}[1]{#1}
\small
\[
\begin{array}{rlcl}
\text{programs} & \rho & ::= & \FF{import}[\Phi](e)\\
\text{fragments} & \Phi & ::= & \emptyset \pipe \Phi, \fvar{tycon}=\{ \delta \}\\
\text{tycon defs} & \delta & ::= & \FF{analit}=\sigma; ~\FF{synidxlit}=\sigma; \\
& & & \FF{anatarg}=\sigma;~\FF{syntarg}=\sigma\\
\text{expressions} & e & ::= & x \pipe \FF{let}(e; x.e) \pipe   \FF{slet}[\sigma](\tvar{x}.e) \pipe \FF{asc}[\sigma](e) \\
& & \pipe & \lambda(x.e) \pipe \FF{lit}[\sigma](\bar{e}) \pipe \FF{targ}[\sigma](e; \overline{e}) \\
\text{static terms} & \sigma & ::= & \tvar{x}  \pipe \noneqval{\lambda(\tvar{x}.\sigma)} \pipe\FF{ap}(\sigma; \sigma) \pipe \FF{fail}\\
 & & \pipe & \eqval{\FF{ty}[\fvar{tycon}](\sigma)} \pipe \FF{tycase}[\fvar{tycon}](\sigma; \tvar{x}.\sigma; \sigma) \\
& & \pipe & \eqval{\FF{incty}[\fvar{tycon}](\sigma)}  \\
& & \pipe & \eqval{\FF{lbl}[{\texttt{lbl}}]} \pipe \FF{lbleq}(\sigma; \sigma; \sigma; \sigma) \\
& & \pipe & \eqval{\FF{nil}} \pipe \eqval{\FF{cons}(\sigma; \sigma)} \pipe  \FF{listrec}(\sigma; \sigma; \tvar{x}.\tvar{y}.\sigma) \\
 & & \pipe & \noneqval{\FF{arg}[e]} \pipe \FF{ana}(\sigma; \sigma; \tvar{x}.\sigma) \pipe \FF{syn}(\sigma; \tvar{x}.\tvar{y}.\sigma)\\
 & & \pipe &  \noneqval{\titerm{\iota}}\\
\text{internal terms} & \iota & ::= & \iup{\sigma} \pipe x \pipe \lambda(x.\iota) \pipe \FF{iap}(\iota; \iota) \\
& & \pipe  &  \FF{inil} \pipe \FF{icons}(\iota; \iota) \pipe \FF{ilistrec}(\iota; \iota; x.y.\iota)

\end{array}
\]
%\vspace{-15px}
\caption{Abstract syntax of @$\lambda$. Metavariable $\fvar{tycon}$ ranges over type constructor names (assumed globally unique), ${\texttt{lbl}}$ over static labels, $x, y$ over expression variables and $\tvar{x}, \tvar{y}$ over static variables. We indicate that variables or static variables are bound within a term or static term by separating them with a dot, e.g. $\tvar{x}.\tvar{y}.e$, and abbreviate a sequence of zero or more expressions as $\overline{e}$. %Static values are shown annotated with either $\color{gray}=$ or $\color{gray}\neq$. Values having and containing only sub-terms with $\color{gray}=$ annotations are those for which syntactic and semantic equality coincide, and only such values are valid type indices, captured in the judgement $\sigma~\mathtt{eqval}_\Phi$.   % Highlighted terms are exclusively used by fragment providers and will be described further below; $\FF{arg}[e]$ is introduced only by the semantics and is a value, and the IL quotation form $\titerm{\iota}$ is a value if it does not contain any unquote forms, $\iup{\sigma}$.
}
\label{syntax-clients}
\vspace{5px}
\end{figure}
%\begin{figure}[t]
%\small
%$\begin{array}{rcl}
%\fvar{tycon}\texttt{[}\sigma\texttt{]} & := & \FF{ty}[\fvar{tycon}](\sigma)\\
%\fvar{tycon}\texttt{[}\cdots\sigma\cdots\texttt{]} & := & \FF{incty}[\fvar{tycon}](\sigma)\end{array}$
%~~~~~~~~$
%\begin{array}{rcl}
%\fvar{tycon} & := & \FF{incty}[\fvar{tycon}](\FF{nil})\\
%\fvar{tycon}\texttt{[}\cdots\texttt{]} & := & \FF{incty}[\fvar{tycon}](\FF{nil})\\
%\end{array}$\\
%~\\
%$
%\begin{array}{rcl}
%\lambda x{:}\sigma.e & := & \FF{asc}[\FF{incty}[\fvar{fn}](\sigma)](\lambda(x.e))\\
%(e_1, ..., e_n) & := & \FF{lit}[\FF{nil}](e_1; \ldots; e_n)\\
%\{{\texttt{lbl}}_1=e_1, \ldots, {\texttt{lbl}}_n=e_n\} & := & \FF{lit}[\FF{cons}(\FF{lbl}[{\texttt{lbl}}_1]; \ldots\FF{cons}(\FF{lbl}[{\texttt{lbl}}_n]; \FF{nil}))](e_1; \ldots; e_n)\\
%\text{(any numeric label)}~~~~{\texttt{n}} & := & \FF{lit}[\FF{lbl}[\texttt{n}]]()\\
%e(e_1) & := & \FF{targ}[\FF{nil}](e; e_1)\\
%e\cdot{\texttt{lbl}} & := & \FF{targ}[\FF{lbl}[{\texttt{lbl}}]](e; \cdot)\\
%e\cdot{\texttt{lbl}}(e_1, ..., e_n) & := & \FF{targ}[\FF{lbl}[{\texttt{lbl}}]](e; e_1; \ldots; e_n)
%%%(e_1, ..., e_n) & := & intro[()](e_1; ...; e_n)
%%%n &  := & intro[n]()
%%%s & := & intro[s]()
%%%
%%%e.l & := & elim[l](e)
%%%e[e_1; ...; e_n] & := & elim[()](e_1, ..., e_n)
%%%e.l(e_1; ...; e_n) & := & elim[l]
%\end{array}
%$
%\vspace{-5px}
%\caption{\small Selected desugarings from a stylized conventional concrete syntax to abstract syntax.}
%\label{desugaring}
%\vspace{10px}
%\end{figure}

%

We now turn our attention to a type theoretic formulation of the key mechanisms described above atop a minimal abstract syntax, shown in Fig. \ref{syntax-clients}. This syntax supports a \emph{purely syntactic desugaring} from a conventional concrete syntax, shown by example in Fig. \ref{atlam-example}.

\paragraph{Fragment Client Perspective} A program, $\rho$, consists of a series of fragment imports, $\Phi$, defining active type constructors for use in an expression, $e$. As described in the introduction, expressions are indexed by static terms, $\sigma$, and indices that can only be trivial are omitted. The abstract syntax of $e$ provides $\FF{let}$ binding of variables and for convenience, static values can also be bound to a static variable, $\tvar{x}$ (distinguished in bold for clarity), using $\FF{slet}$. Static terms have a \emph{static dynamics} and evaluate to \emph{static values} or fail.

\subparagraph{Types and Ascriptions} An expression can be ascribed a {type} or an {incomplete type}, both  {static values} constructed, as in the introduction, by naming an imported type constructor, $\fvar{tycon}$, and providing a type index, another static value. The static language also includes lists and atomic \emph{labels} for use in compound indices. %The remaining static term constructors are of use mainly to fragment providers (see below).

\subparagraph{Literal Desugaring} All concrete literals (other than lambda expressions, which are built in)  desugar to an abstract term of the form $\FF{lit}[\sigma](\overline{e})$, where $\sigma$ captures all static portions of the literal (e.g. a list of the labels used as field names in the labeled product literal in Fig. \ref{atlam-example}, or the numeric label used for the natural number zero) and $\overline{e}$ is a list of sub-expressions (e.g. the field values). % is only one term constructor corresponding to each basic method by which a type constructor is delegated responsibility over a term.

\subparagraph{Targeted Expression Desugaring} All non-introductory operations go through a targeted expression form (e.g. $e(e)$, or $e\cdot\texttt{lbl}$, or $e\cdot\texttt{lbl}(\overline{e})$; see previous section), which all desugar to an abstract term of the form $\FF{targ}[\sigma](e; \overline{e})$ where $\sigma$ again captures all static portions of the term (e.g. the label naming an operation, e.g. $\texttt{s}$ or $\texttt{rec}$ on natural numbers \cite{pfpl}), $e$ is the  target (e.g. the natural number being operated on, the function being called, or the record we wish to project out of) and $\overline{e}$ are all other arguments (e.g. the base and recursive cases of the recursor, or the argument being applied).% We conjecture that formalizing these semantics would not present a substantial challenge.

\paragraph{Bidirectional Active Typechecking and Translation}
\newcommand{\atjsynX}[3]{\Gamma \vdash_\fvalCtx #1 \Rightarrow #2 \leadsto #3}
\newcommand{\atjanaX}[3]{\Gamma \vdash_\fvalCtx #1 \Leftarrow #2 \leadsto #3}
\newcommand{\atjerrX}[1]{\Gamma \vdash_\fvalCtx #1~ \mathtt{error}}
\begin{figure*}[t!]
\small
$\fbox{\inferrule{}{\atjsynX{e}{\sigma}{\iota}}}$~~
$\fbox{\inferrule{}{\atjanaX{e}{\sigma}{\iota}}}$~~
$\Gamma ::= \emptyset \pipe \Gamma, x \Rightarrow \sigma$
%$\fbox{\inferrule{}{\atjerrX{e}}}$
\renewcommand{\MathparLineskip}{\lineskiplimit=.6\baselineskip\lineskip=.6\baselineskip plus .2\baselineskip}
\begin{mathpar}
\inferrule[att-subsume]{
  \atjsynX{e}{\sigma}{\iota}
}{
  \atjanaX{e}{\sigma}{\iota}
}

\inferrule[att-var]{
  x \Rightarrow \sigma \in \Gamma
}{
  \atjsynX{x}{\sigma}{x}
}

\inferrule[att-ana-let]{
  \atjsynX{e_1}{\sigma_1}{\iota_1}\\\\
  \Gamma, x \Rightarrow \sigma_1 \vdash_\fvalCtx e_2 \Leftarrow \sigma_2 \leadsto \iota_2
}{
  \atjanaX{\FF{let}(e_1; x.e_2)}{\sigma_2}{\FF{iap}(\lambda(x.\iota_2); \iota_1)}
}

\inferrule[att-syn-let]{
  \atjsynX{e_1}{\sigma_1}{\iota_1}\\\\
  \Gamma, x \Rightarrow \sigma_1 \vdash_\fvalCtx e_2 \Rightarrow \sigma_2 \leadsto \iota_2
}{
  \atjsynX{\FF{let}(e_1; x.e_2)}{\sigma_2}{\FF{iap}(\lambda(x.\iota_2); \iota_1)}
}

\inferrule[att-syn-asc]{
    \sigma \Downarrow_{\emptyset; \emptyset} \sigma'\\
     \sigma'~\mathtt{ty}_\fvalCtx\\
  \atjanaX{e}{\sigma'}{\iota}
}{
  \atjsynX{\FF{asc}[\sigma](e)}{\sigma'}{\iota}
}

\inferrule[att-ana-slet]{
   \sigma_1 \Downarrow_{\emptyset; \emptyset} \sigma_1'\\
    \atjanaX{[\sigma_1'/\tvar{x}]e}{\sigma_2}{\iota}
}{
  \atjanaX{\FF{slet}[\sigma_1](\tvar{x}.e)}{\sigma_2}{\iota}
}

\inferrule[att-syn-slet]{
   \sigma_1 \Downarrow_{\emptyset; \emptyset} \sigma_1'\\
    \atjsynX{[\sigma_1'/\tvar{x}]e}{\sigma_2}{\iota}
}{
  \atjsynX{\FF{slet}[\sigma_1](\tvar{x}.e)}{\sigma_2}{\iota}
}

\inferrule[att-ana-lam]{
    \sigma_1~\mathtt{ty}_\fvalCtx\\\sigma_2~\mathtt{ty}_\fvalCtx\\
  \Gamma, x \Rightarrow \sigma_1 \vdash_\fvalCtx e \Leftarrow \sigma_2 \leadsto \iota
}{
  \atjanaX{\lambda x.e}{\FF{ty}[\fvar{fn}](\FF{cons}(\sigma_1; \FF{cons}(\sigma_2; \FF{nil})))}{\lambda(x.\iota)}
}

\inferrule[att-syn-idx-lam]{
    \sigma_1~\mathtt{ty}_\fvalCtx\\
  \Gamma, x \Rightarrow \sigma_1 \vdash_\fvalCtx e \Rightarrow \sigma_2 \leadsto \iota
}{
  \atjsynX{\FF{asc}[\FF{incty}[\fvar{fn}](\sigma_1)](\lambda(x.e))}{\FF{ty}[\fvar{fn}](\FF{cons}(\sigma_1; \FF{cons}(\sigma_2; \FF{nil})))}{\lambda(x.\iota)}
}
%
%\inferrule[att-lam]{
% \tau_1 \Downarrow_{\Gamma;\fvalCtx} \tau_1'\\
%   \trepof{\tau_1'} \Downarrow_{\Gamma;\fvalCtx} \titype{\sigma}\\\\
% \Gamma, x \Rightarrow \tau_1' \vdash_\fvalCtx e \Rightarrow \tau_2 \leadsto \iota
%%  \iota \hookrightarrow_\fvalCtx \iota'\\
%%    \sigma \hookrightarrow_\fvalCtx \sigma'
%%  \ddbar{\fvar{Arrow}}{\fvalCtx}{\trepof{\tau_1'}}{\sbar_1}\\
% %\delfromtau{$\Xi_0$}{\fvalCtx}{\tau_1'}{\sabs}\\\\
%}{
% \atjsynX{\elam{x}{\tau_1}{e}}{\ttype{arrow}{(\tau_1', \tau_2)}}{\ilam{x}{\sigma'}{\iota}}
%}

\inferrule[att-ana-lit]{
  \vdash_\fvalCtx \FF{analit}(\fvar{tycon})=\sigmat{def}\\
  \mathtt{args}(\overline{e})=\sigmat{args}\\\\
  \sigmat{def} ~\sigmat{tyidx} ~\sigmat{tmidx} ~\sigmat{args} \Downarrow_{\Gamma; \fvalCtx} \titerm{\iota}
% \FF{ap}(\FF{ap}(\FF{ap}(\sigmat{def}; \sigmat{tyidx}); \sigmat{tmidx}); \sigmat{args}) \Downarrow_{\Gamma;\fvalCtx} \titerm{\iota}
% \trepof{\ttype{tycon}{\tauidx'}} \Downarrow_{\Gamma;\fvalCtx} \titype{\sigma}\\
}{
  \atjanaX{\FF{lit}[\sigmat{tmidx}](\overline{e})}{\FF{ty}[\fvar{tycon}](\sigmat{tyidx})}{\iota}
}

\inferrule[att-syn-idx-lit]{
  \vdash_\fvalCtx \FF{synidxlit}(\fvar{tycon})=\sigmat{def}\\
  \mathtt{args}(\overline{e})=\sigmat{args}\\\\
  \sigmat{def} ~\sigmat{incidx} ~\sigmat{tmidx} ~\sigmat{args} \Downarrow_{\Gamma;\fvalCtx} \FF{cons}(\sigmat{tyidx}; \FF{cons}(\titerm{\iota}; \FF{nil}))\\
% \FF{ty}[\fvar{tycon}](\sigmat{tyidx})~\mathtt{ty}_\fvalCtx
% \FF{ap}(\FF{ap}(\FF{ap}(\sigmat{def}; \sigmat{incidx}); \sigmat{tmidx}); \sigmat{args}) \Downarrow_{\Gamma;\fvalCtx}\FF{cons}(\sigmat{tyidx}; \FF{cons}(\titerm{\iota}; \FF{nil}))
}{
  \atjsynX{\FF{asc}[\FF{incty}[\fvar{tycon}](\sigmat{incidx})](\FF{lit}[\sigmat{tmidx}](\overline{e}))}{\FF{ty}[\fvar{tycon}](\sigmat{tyidx})}{\iota}
}
%
%\inferrule[att-i-asc-ty]{
% \atjanaX{I}{\tau}{\iota}
%}{
% \atjsynX{I : \tau}{\tau}{\iota}
%}
%
%\inferrule[att-i-asc-tycon]{
% \vdash_\fvalCtx \FF{isyn}(\fvar{tycon})=\taudef\\\\
% \taudef~\tauidx~((\Gamma; e_1)? {::}{\ldots}{::}(\Gamma; e_n)? {::} []) \Downarrow_{\Gamma;\fvalCtx} \tden{\titerm{\iota}}{\ttype{tycon}{\tauidx'}}\\
%     \trepof{\tau_1'} \Downarrow_{\Gamma;\fvalCtx} \titype{\sigma}\\
% \Gamma \vdash_\fvalCtx \iota : \sigma
%}{
% \atjsynX{\FF{intro}[\tauidx](e_1; \ldots; e_n)] :: \fvar{tycon}}{\ttype{tycon}{\tauidx'}}{\iota'}
%}

\inferrule[att-ana-targ]{
  \atjsynX{e_\text{targ}}{\FF{ty}[\fvar{tycon}](\sigmat{tyidx})}{\iota_\text{targ}}\\\\
  \vdash_\fvalCtx \FF{anatarg}(\fvar{tycon})=\sigmat{def}\\
  \mathtt{args}(\overline{e})=\sigmat{args}\\\\
  \sigmat{def} ~\sigmat{tyidx} ~\titerm{\iota_\text{targ}} ~\sigmat{ty} ~\sigmat{tmidx} ~\sigmat{args} \Downarrow_{\Gamma;\fvalCtx} \titerm{\iota}
% \FF{ap}(\FF{ap}(\FF{ap}(\FF{ap}(\FF{ap}(\sigmat{def}; \sigmat{tyidx}); \titerm{\iota}); \sigma); \sigmat{tmidx}); \sigmat{args}) \Downarrow_{\Gamma;\fvalCtx} \titerm{\iota'}
}{
  \atjanaX{\FF{targ}[\sigmat{tmidx}](e_\text{targ}; \overline{e})}{\sigmat{ty}}{\iota}
}

\inferrule[att-syn-targ]{
  \atjsynX{e_\text{targ}}{\FF{ty}[\fvar{tycon}](\sigmat{tyidx})}{\iota_\text{targ}}\\\\
  \vdash_\fvalCtx \FF{syntarg}(\fvar{tycon})=\sigmat{def}\\
  \mathtt{args}(\overline{e})=\sigmat{args}\\\\
  \sigmat{def} ~\sigmat{tyidx} ~\titerm{\iota_\text{targ}} ~\sigmat{tmidx} ~\sigmat{args} \Downarrow_{\Gamma;\fvalCtx} \FF{cons}(\sigmat{ty}; \FF{cons}(\titerm{\iota}; \FF{nil}))\\
  \sigmat{ty}~\mathtt{ty}_\fvalCtx
}{
  \atjsynX{\FF{targ}[\sigmat{tmidx}](e_\text{targ}; \overline{e})}{\sigmat{ty}}{\iota'}
}
\end{mathpar}
\caption{Bidirectional active typechecking and translation. For concision, we use standard functional notation for static function application.}
\label{atj}
\end{figure*}
%In this section, we will describe the foundations of \emph{active typechecking and translation} by constructing a core lambda calculus, @$\lambda$. 
%A \emph{program} in @$\lambda$, $\rho$, consists of a series of fragment definitions, $\phi$, for use by an external term, $e$. An example of an @$\lambda$ program, $\rho_{\text{example}}$, that uses {fragment definitions} $\phi_{\texttt{nat}}$ and $\phi_{\texttt{lprod}}$, defining primitive natural numbers and labeled products (i.e. ordered records), is shown in Fig. \ref{nat-sugared}. In @\texttt{lang},  fragment definitions are packaged separately; we do not include this in the core calculus for simplicity.
%
%In this example, we see the introduction and elimination forms for natural numbers, functions and labeled tuples all being used. However, in the core syntax, shown (from the perspective of fragment clients) in Fig. \ref{client-syntax}, there are only two external introductory forms, $\lambda x.e$ and $\FF{intro}[\sigma](\overline{e})$, and one external elimination form, $e\cdot\FF{elim}[\sigma](\overline{e})$, where $\overline{e}$ is shorthand for zero or more semicolon-separated external terms, called the \emph{arguments} of the introduction or elimination operation. The forms used in Fig. \ref{nat-sugared} are \emph{derived forms}, defined in terms of these core forms in Fig. \ref{derived}. 
%
%Both fragments and external terms can contain \emph{static terms}, $\sigma$, which are evaluated statically to \emph{static values}, $\hat\sigma$. The static language is itself a simply-typed lambda calculus. We call the types of static terms \emph{kinds}, $\kappa$, by convention. Static variables are written in bold font, $\tvar{x}$, to emphasize that they are distinct from variables bound by external terms, written $x$. For example, in the example in Figure \ref{nat-sugared}, $\tvar{nat}$ is a static variable bound by the fragment $\phi_{\texttt{nat}}$ to the natural number type. Types have kind $\kTypeBlur$ and are introduced by naming a \emph{type constructor}, written in small-caps, e.g. $\fvar{nat}$ and $\fvar{lprod}$, and providing an \emph{index}, which is a static term of the \emph{index kind} of the type constructor.  For example, $\fvar{nat}$ is indexed by $\kunit$ (because there is only one natural number type), so $\tvar{nat}$ is defined to be $\ttype{nat}{\tunit}$, and $\fvar{lprod}$ is indexed by $\klist{\kpair{\FF{L}}{\kTypeBlur}}$, classifying static lists pairing \emph{labels} with types. Labels are static values of kind $\FF{L}$, written abstractly using the metavariables containing $\ell$ in our examples. 
%
%Before typechecking the external term in a program, the program is kind checked and its static terms are normalized to static values.  We write static values using the hatted metavariable $\hat \sigma$. Other syntactic metavariables in our system also have hatted forms if they can contain static terms. The desugared, statically normalized version of the example in Figure \ref{nat-sugared}, $\hat\rho_{\text{example}}$, is shown in Figure \ref{nat-normal}. %The syntax of @$\lambda$ from the perspective of a fragment client is given in Fig. \ref{atlam-syntax}. We will see a few additions relevant only to fragment providers when we discuss defining fragments below.
%
%\subsection{External Terms}
% a program consists of first \emph{kind checking} its static terms, then normalizing its static terms, then typechecking the external term and,  simultaneously, \emph{translating} it to a term, $\iota$, in the \emph{typed internal language}. The dynamic behavior of an external term is determined entirely by its translation to this language. Internal types are written $\tau$. The internal language is only exposed to fragment providers, not to clients (i.e. normal developers).

%We will now give a core typed lambda calculus, @$\lambda$, that captures the semantics described in the previous sections. It is intended to make precise how active typechecking and translation works and how our mechanism relates to existing work on bidirectional typechecking, type-level computation and typed compilation while abstracting away from the details of Python's syntax and imposing a stronger type-level semantics that will allow us to state metatheoretic properties of interest. We will assume a fixed base for functions (providing the standard semantics of lambda functions) and target language, which we here call the \emph{internal language}.
 % program written in core @$\lambda$, without syntactic sugar and with all static terms normalized, is shown in Fig. \ref{nat-desugared}. The syntax of core @$\lambda$ is shown in Fig. \ref{grammar} and the syntactic desugarings are shown in Fig. \ref{desugaring}. The definition of $\phi_{\texttt{nat}}$ is shown in Fig. \ref{nat-atfrag}. We will discuss these in the following sections.

%In this section, we will develop an ``actively typed'' version of the simply-typed lambda calculus with simply-kinded type-level computation called $\lam@\texttt{lang}$. More specifically, the level of types, $\tau$, will itself form a simply-typed lambda calculus. \emph{Kinds} classify type-level terms in the same way that types conventionally classify expressions. Types become just one kind  of type-level value (which we will write $\kTypeBlur$, though it is also variously written $\star$, \lstinlinep{T} and \lstinlinep{Type} in various settings). Rather than there being a fixed set of type and operator constructors, we allow the programmer to declare new  constructors, and give their static and dynamic semantics by writing type-level functions. The kind system combined with techniques borrowed from the typed compilation literature and a form of type abstraction will allow us to prove strong type safety, decidability and conservativity theorems.
%
%The syntax of Core $\lam@\texttt{lang}$ is given in Fig. \ref{grammar}. An example of a program defining type and operator constructors that can be used to construct an active embedding of G\"odel's \textbf{T} into $\lam@\texttt{lang}$ is given in Fig. \ref{nat}. We will discuss its semantics and how precisely the embedding, seen being used starting on line 15 to ultimately compute the sum of two and two, works as we go on. Natural numbers can, of course, be isomorphically embedded in existing languages, with a similar usage and asymptotic performance profile (up to function call overhead as an abstract type, for example). We will provide more sophisticated examples where this is less feasible later on (and note that type abstraction is an orthogonal mechanism).
%a type constructor declaration, $\fvar{nat}$, indexed trivially, together with three operator constructors, also all indexed trivially, that implement the standard introductory forms for natural numbers as well as the recursor operator (as in G\"odel's T \cite{pfpl}). Following the type constructor declaration, we apply $\fvar{nat}$ with the trivial index, $\tunit$, to form the type $\tvar{nat}$. Finally, we write an external term that uses the operators associated with $\fvar{nat}$ and the built-in constructor $\fvar{Parr}$, governing partial functions, to define an addition function and compute the addition of the natural numbers  two and two. We will introduce a more convenient concrete syntax in later portions of this thesis; for now we will restrict ourselves to the abstract syntax so that this example can directly aid in understanding the semantics.

%\begin{figure}[t]
%\small
%$$\begin{array}{rrcl}  
%%\textbf{programs} & 
%\text{programs} & \rho & ::= & \F{using}\phi~\F{in}e \\
%\text{fragments} & \phi &  ::= & \F{tycon}\fvar{tycon}~\F{of}\kappat{tyidx}~{\{}\F{iana} \tau_1; \F{esyn} \tau_2; \F{rep} \tau_3{\}} %\pipe \phi; \phi %\\
%%\\
%%& \pipe & 
%\pipe \F{def}\tvar{t} : \kappa = \tau \pipe \phi; \phi%\pdef{t}{\kappa}{\tau}{\progsort} \pipe 
%\\ %\pfam{\familyDf}{\progsort}
%%\\&  \pipe & \pdef{t}{\kappa}{\tau}{\progsort}  \pipe e
%%\pipe \pdef{t}{\kappa}{\tau}{\progsort} 
%\text{external terms}& e & ::= & \evar{x} \pipe \F{let} x = e_1~\F{in} e_2 \pipe \lambda x.e \pipe e : \tau\\
%&& \pipe & \FF{intro}[\taut{opidx}](\splat{e}{1}{n})\pipe e\cdot\FF{elim}[\taut{opidx}](\splat{e}{1}{n}) \\
%
%\text{type-level terms} & \tau   & ::=   &   \tvar{t} \pipe \tifeq{\tau_{1}}{\tau_{2}}{\kappa}{\tau_{3}}{\tau_{4}} \pipe \terr  \\
% && \pipe & 
% %\F{let}\tvar{t}{:}\kappa=\tau_1~\F{in} \tau_2 \pipe 
%                           \tlam{t}{\kappa}{\tau} \pipe 
%                           \tapp{\tau_1}{\tau_2} \\
%&&\pipe&                     
%                           \tnil{\kappa} \pipe \tcons{\tau_1}{\tau_2} \pipe 
%                                      \tfold{\tau_1}{\tau_2}{\tvar{t}_{hd}}{\tvar{t}_{tl}}{\tvar{t}_{rec}}{\tau_3}
%                           \\
%     && \pipe  &     \ell  \\
%     && \pipe &  
%   \tunit \pipe 
%                           \tpair{\tau_{1}}{\tau_{2}} \pipe 
%                           \tfst{\tau} \pipe 
%                           \tsnd{\tau} 
%                           \\  
%     && \pipe & \tinl{\kappa_1, \kappa_2}{\tau_1} \pipe \tinr{\kappa_1, \kappa_2}{\tau_2} \pipe \tsumcase{\tau}{t}{\tau_1}{t}{\tau_2}\\
%&&\pipe  &   \ttype{tycon}{\taut{tyidx}} \pipe  \tfamcase{\tau}{tycon}{t}{\tau_1}{\tau_2}\\
%%                            \\                          %       & & \pipe & \tfamcase{\tau}{Fam}{x}{\tau_1}{\tau_2}\\
%%                                                
%%\\     &  \pipe & \tden{\tau_2}{\tau_1} \pipe \terr %\pipe \tden{\ibar}{\tau}^{\checkmark} 
%%\pipe \ttypeof{\tau} \pipe \itransof{\tau} \\% \tdencase{\tau}{x}{t}{\tau_1}{\tau_2}\\
%% %& & \pipe & 
%%%                           \tdencase{\tau}{y}{x}{\tau_1}{\tau_2}
%%%                            \\
%%
% && \pipe & \titerm{\iota}
%     \pipe \titype{\sigma} \pipe \trepof{\tau} \\
%   && \pipe &  (\Gamma; e)? \pipe \FF{syn}(\tau_1; \tvar{t}_{ty},\tvar{t}_{trans}.\tau_2) \pipe \FF{ana}(\tau_1; \tau_2; \tvar{t}_{trans}.\tau_3)\\
%%\textbf{external terms}         & e & ::= & \evar{x} \pipe 
%%%                           \efix{x}{\tau}{e} \pipe 
%%                            \elam{\evar{x}}{\tau}{e} \pipe 
%%                            \eop{tycon}{op}{
%%                              \tauidx
%%                            }{
%%                                  \splat{e}{1}{n}
%%                            } \\
%%                  &     &   &   \\
%%
%%
%%                        
%%%\text{deabstracted}& \iota & ::= & \mathcal{G}[\iota, \sigma]\\
%%                        
%%\\
%%\\
%%              
%%% &  & \pipe & \tvalof{\tau_1}{\tau_2} \pipe \iup{\tau} \\
%%%                         \trepof{\tau} \pipe \dup{\tau}\\
%%\text{translational IL} & \bar{\iota} & ::= & x \pipe \ifix{x}{\bar \sigma}{\bar \iota} 
%%  %\pipe \ilam{x}{\bar \sigma}{\bar \iota} \pipe \iapp{\bar \iota_1}{\bar \iota_2} 
%%  \pipe \cdots \pipe \itransof{\tau} \\           
%% & \bar{\sigma} & ::= & \darrow{\bar \sigma_1}{\bar \sigma_2} \pipe \cdots \pipe \dup{\tau} \pipe \trepof{\tau} \\
%%%\text{abstracted} & \sabs & ::= & \darrow{\sabs_1}{\sabs_2} \pipe \cdots \pipe \sabsrep{\tau}
%%                      \\
%\text{kinds} & \kappa  & ::= & \karrow{\kappa_1}{\kappa_2} \pipe \klist{\kappa} \pipe \FF{L} \pipe \kunit \pipe \kpair{\kappa_1}{\kappa_2} \pipe \ksum{\kappa_1}{\kappa_2} \pipe \kTypeBlur \pipe \kITerm \pipe \kIType \pipe \Q
%%                          \klabel \pipe
%%                          \klist{\kappa} \pipe
%%                        \kunit \pipe 
%%                        \kpair{\kappa_{1}}{\kappa_{2}} \\
%%                        &&\pipe&
%%                        \ksum{\kappa_1}{\kappa_2} \pipe
%%                        \kTypeBlur \pipe \kDen \pipe 
%%                        \kIType               
%%%\textbf{ops signature}     & \Theta  & ::= & \kOpEmpty \pipe \kOp{\Theta}{op}{\kappai}\\
%%%                                   &   &   & \\
%\\
%\text{typing contexts} & \Gamma & ::= & \emptyset \pipe \Gamma, x \Rightarrow \tau\\
%\\
%\text{internal terms} & \iota  & ::= & \evar{x} \pipe 
%                       \ifix{\evar{x}}{\sigma}{\iota} \pipe
%                       \ilam{\evar{x}}{\sigma}{\iota} \pipe 
%                       \iapp{\iota_{1}}{\iota_{2}} \pipe n \pipe \iota_1 \pm \iota_2 \pipe \FF{if0}(\iota; \iota_1; \iota_2)\\
%   && \pipe &  \iunit \pipe \ipair{\iota_1}{\iota_2} \pipe \ifst{\iota} \pipe \isnd{\iota} \pipe \iup{\tau} 
%%\\&\pipe&               \cdots \pipe  
%%                        n \pipe \iop{\iota_{1}}{\iota_{2}} \pipe \iIfEq{\iota_{1}}{\iota_{2}}{\dint}{\iota_{3}}{\iota_{4}} \\
%%& \pipe & 
%%                        \iunit \pipe
%%                        \ipair{\iota_{1}}{\iota_{2}} \pipe 
%%                        \ifst{\iota} \pipe
%%                        \isnd{\iota} 
%%\\&\pipe&                       
%%                         \iinl{\sigma_2}{\iota_1} \pipe \iinr{\sigma_1}{\iota_2} \\&\pipe & \icase{e}{x}{e_1}{x}{e_2} \\
%\\ \text{internal types} & \sigma  & ::= &    \darrow{\sigma_1}{\sigma_2} \pipe \dint \pipe \dunit \pipe \dpair{\sigma_1}{\sigma_2} %\pipe \dint \pipe \dunit \pipe \dpair{\sigma_1}{\sigma_2} \pipe \dsum{\sigma_1}{\sigma_2}  
% \pipe \dup{\tau}
%\end{array}$$
%%\vspace{-10pt}
%\caption{\small Syntax of Core @$\lambda$. Here, $x$ ranges over external and internal language variables, $\tvar{t}$ ranges over type-level variables, $\fvar{tycon}$ ranges over type constructor names, $\ell$ ranges over labels, $n$ ranges over integers and $\pm$ ranges over standard binary operations over integers. The introductory form for arguments, $(\Gamma; e)?$, should only be constructed by the compiler, not by type constructor providers.
%\label{grammar}}
%\end{figure}
%\begin{figure*}[t]
%\small
%\begin{tabular}{r l l l}
% & \textbf{initial} & \textbf{statically normal}\\
%\textbf{programs} & $\rho ::=$ & $\hat\rho ::=$\\
% & ~~$\FF{using}~{\phi}~\FF{in}~e$ & ~~$\FF{using}~{\hat\phi}~\FF{in}~\hat e$\\
%\textbf{fragment decls.} & $\phi ::=$ & $\hat\phi ::=$\\
%tycon decls. & ~~$\FF{tycon}~\fvar{tycon} : \Theta = \theta$ & ~~$\FF{tycon}~\fvar{tycon} : \Theta = \hat\theta$\\
%static binding & ~~$\FF{def}~\tvar{x} : \kappa = \sigma$ & \\
% & ~~$\phi; \phi$ & ~~$\hat\phi; \hat\phi$\\
%\textbf{static terms} & $\sigma ::=$ & $\hat\sigma ::= $ &\textbf{kinds}~~~  $\kappa ::= $\\
%static values & ~~$\hat \sigma$\\
%static variables & ~~$\tvar{x}$ & ~ & ~\\
%types & ~~$\ttype{tycon}{\sigma}$ & ~~$\ttype{tycon}{\hat\sigma}$ & ~~$\kTypeBlur$\\
%%static products & ~~$\tunit$ & ~~$\tunit$ & ~~$\kunit$\\
%%& ~~$(\sigma, \sigma)$ & ~~$(\hat\sigma, \hat\sigma)$ & ~~$\kpair{\kappa}{\kappa}$\\
%%static sums & ~~$\FF{in}[\ell](\sigma)$ & ~~$\FF{in}[\ell](\hat\sigma)$ & ~~$\Sigma[\{\overline{\ell \hookrightarrow \kappa}\}]$\\
%static lists & ~~$[]_\kappa$ & ~~$[]_\kappa$ & ~~$\klist{\kappa}$\\
% & ~~$\sigma :: \sigma$ & ~~$\hat\sigma :: \hat\sigma$ & \\
%static labels & ~~$\ell$ & ~~$\ell$ & ~~$\FF{L}$\\
%\\
%\textbf{external terms} & $e ::= $ & $\hat e ::=$ \\
%variables & ~~$x$ & ~~$x$\\
%binding & ~~$\FF{let}~x=e~\FF{in}~e$ & ~~$\FF{let}~x=\hat{e}~\FF{in}~\hat{e}$\\
%fixpoints & ~~$\FF{fix}~x{:}\sigma~\FF{is}~e$ & ~~$\FF{fix}~x{:}\hat\sigma~\FF{is}~\hat e$\\
%ascription & ~~$e : \sigma$ & ~~$\hat{e} : \hat\sigma$\\
%%introductory forms & ~~$\nu$ & ~~$\hat\nu$\\
%%eliminatory forms & ~~$\eta$ & ~~$\hat\eta$\\
%% & $\nu ::= $ & $\hat \nu ::=$\\
%analytic lambda & ~~$\lambda x.e$ & ~~$\lambda x.\hat{e}$\\
%%synthetic lambda & ~~$\lambda x {:} \sigma.e$ & ~~$\lambda x{:}\hat\sigma.\hat e$\\
%analytic intro & ~~$\FF{intro}[\sigma](\overline{e})$ & ~~$\FF{intro}[\hat\sigma](\overline{\hat e})$\\
%%synthetic intro & ~~$\fvar{tycon}\cdot\FF{intro}[\sigma](\overline{e})$ & ~~$\fvar{tycon}\cdot\FF{intro}[\hat \sigma](\overline{\hat e})$\\
%% & $\eta ::= $ & $\hat\eta ::=$\\
%synthetic elim & ~~$e\cdot\FF{elim}[\sigma](\overline{e})$ & ~~$\hat e\cdot\FF{elim}[\hat \sigma](\overline{\hat e})$\\
%\end{tabular}
%\caption{Syntax from the perspective of fragment clients (i.e. normal developers)}
%\end{figure*}
%\begin{figure*}[t]
%\small
%\hspace{-5px}
%$
%\begin{array}{rclr}
%%\FF{intro}(e_1; \ldots; e_n) & := & \FF{intro}[()](e_1; \ldots; e_n)\\
%\sigma\langle e_1; \ldots; e_n \rangle & := & \FF{intro}[\tfst{\sigma}](e_1; \ldots; e_n) : \tsnd{\sigma} & \text{assisted intro}\\
%%\langle\sigmaidx\rangle(e_1; \ldots; e_n) & := & \FF{intro}[\sigmaidx](e_1; \ldots; e_n)\\
%%[\sigmaidx](e_1; \ldots; e_n) & := & \FF{intro}[\sigmaidx](e_1; \ldots; e_n)\\
%\{ \ell_1{=}~e_1, \ldots, \ell_n{=}~e_n \} & := & \FF{intro}[\ell_1 :: \ldots :: \ell_n :: []_{\FF{L}}](e_1; \ldots; e_n) & \text{labeled seq.}  \\
%(e_1, ..., e_n) & := & \FF{intro}[()](e_1; \ldots; e_n) & \text{unlabeled seq.}\\
%%n & := & \FF{intro}[n]() & \text{number literal} \\
%%s & := & \FF{intro}[s]() & \text{string literal}\\
%%e\cdot\FF{elim}(e_1; \ldots; e_n) & := & e\cdot\FF{elim}[()](e_1; \ldots; e_n)\\
%e~e_1 & := & e\cdot\FF{elim}[()](e_1) & \text{application}\\
%\sigma~e~\langle e_1; \ldots; e_n \rangle  & := & e\cdot\FF{elim}[\sigma](e_1; \ldots; e_n) & \text{assisted elim}\\
%e.\ell & := & e\cdot\FF{elim}[\ell]() & \text{projection}\\
%%\F{case}e~\{\ell_1\langle x_1 \rangle \Rightarrow e_1~|~\ldots~|~\ell_n\langle x_n \rangle \Rightarrow e_n\} & := & e\cdot\FF{elim}[\ell_1 :: \ldots :: \ell_n :: []_{\FF{L}}]( & \text{case analysis}\\
%% & & \quad \lambda x_1.e_1; \ldots; \lambda x_n.e_n)\\
%\sigma_1 \rightharpoonup \sigma_2 & := & \ttype{parr}{(\sigma_1, \sigma_2)} & \text{arrow types}
%%!TEX encoding = UTF-8 Unicode
%\begin{figure*}
%\newcommand{\tcl}[1]{\multicolumn{2}{l}{#1}}
%\small
%\begin{tabular}{r l l l l}
% & \tcl{\textbf{initial}} & \textbf{statically normal} \\
%\textbf{tycon definitions} & \tcl{$\delta ::=$}  & $\hat\delta ::=$ \\
% & \tcl{~~$\FF{iana}~\sigma; \FF{esyn}~\sigma; \FF{rep}~\sigma$} & ~~$\FF{iana}~\hat\sigma; \FF{esyn}~\hat\sigma; \FF{rep}~\hat\sigma$\\
% \\
%\textbf{(static language, cont.)} & \tcl{$\sigma ::= ...$} & $\hat\sigma ::= ...$ & $\kappa ::= ...$\\
%%type elim & \tcl{~~$\tfamcase{\sigma}{tycon}{x}{\sigma}{\sigma}$} & \\
%%static product elim & \tcl{~~$\sigma.\FF{prl}$} &\\
%% & \tcl{~~$\sigma.\FF{prr}$} &\\
%%static sum elim & \tcl{~~$\FF{case}~\sigma~\{\overline{l\langle\tvar{x}\rangle\Rightarrow\sigma}\}$} &\\
%%static list elim & \tcl{~~$\FF{listrec}(\sigma; \sigma; \tvar{h}, \tvar{t}, \tvar{r}.\sigma)$} & ~\\
%decidable equality & \tcl{~~$\tifeq{\hat\sigma}{\hat\sigma}{\kappa}{\hat\sigma}{\hat\sigma}$} & ~ & ~\\
%type error & \tcl{~~$\FF{tyerr}$} & ~ \\
%%static functions & \tcl{~~$\tlam{x}{\kappa}{\sigma}$} & ~~$\tlam{x}{\kappa}{\sigma}$ & ~~$\karrow{\kappa}{\kappa}$\\
%% & \tcl{~~$\sigma~\sigma$} & ~ & ~\\
%term translations & \tcl{~~$\titerm{\tri}$} & ~~$\titerm{\iota}$ & ~~$\kITerm$\\
%type translations & \tcl{~~$\titype{\trt}$} & ~~$\titype{\tau}$ & ~~$\kIType$\\
% & \tcl{~~$\trepof{\sigma}$} & ~\\
%arguments & \multicolumn{3}{c}{\color{gray}(introduced only by typechecker)} & ~~$\Q$\\
% & \tcl{~~$\FF{ana}(\sigma; \sigma; \tvar{x}.\sigma)$} & \\
% & \tcl{~~$\FF{syn}(\sigma; \tvar{x}, \tvar{t}.\sigma)$} & \\
%\\
%\textbf{internal language} & $\tri ::= $ & $\trt ::= $ & $\iota ::= $ & $\tau ::=$\\
%splice & ~~$\iup{\sigma}$ & ~~$\dup{\sigma}$ & \\
%variables & ~~$x$ &  & ~~$x$ & \\
%%fixpoints & ~~$\FF{fix}~x{:}\trt~\FF{is}~\tri$ & & ~~$\FF{fix}~x{:}\tau~\FF{is}~\iota$\\
%%functions & ~~$\lambda x{:}\trt.\tri$ & ~~$\trt \rightharpoonup \trt$ & ~~$\lambda x{:}\tau.\iota$ & ~~$\tau \rightharpoonup \tau$\\
%% & ~~$\tri~\tri$ & ~ & ~~$\iota~\iota$\\
%%products & ~~$()$ & ~~$\dunit$ & ~~$()$ & ~~$\dunit$\\
%% & ~~$(\tri, \tri)$ & ~~$\dpair{\trt}{\trt}$ & ~~$(\iota, \iota)$ & ~~$\dpair{\tau}{\tau}$\\
%% & ~~$\tri.\FF{prl}$ & ~ & ~~$\tri.\FF{prl}$ & ~\\
%%  & ~~$\tri.\FF{prr}$ & ~ & ~~$\tri.\FF{prr}$ & ~\\
%%%products & ~~$(\overline{l=\tri})$ & ~~$\Pi[\overline{(l, \trt)}]$ & ~~$(\overline{l = \iota})$ & ~~$\Pi[\overline{(l, \tau)}]$\\
%%% & ~~$\tri\cdot l$ & ~ & ~~$\iota\cdot l$ & ~\\
%%sums & ~~$\FF{in}[\ell](\tri)$ & ~~$\Sigma[\{\overline{\ell \hookrightarrow \trt}\}]$ & ~~$\FF{in}[\ell](\iota)$ & ~~$\Sigma[\{\overline{\ell \hookrightarrow \tau}\}]$\\
%% & ~~$\FF{case}~ \tri ~\{\overline{\ell(x) \Rightarrow \tri}\}$ & ~ & ~~$\FF{case}~\iota~\{\overline{\ell(x)\Rightarrow \iota}\}$ & ~\\
%%integers & ~~$\FF{int}[z]$ & ~~$\FF{int}$ & ~~$\FF{int}[z]$ & ~~$\FF{int}$\\
%% & ~~$\tri \pm \tri$ & ~ & ~~$\iota \pm \iota$ & ~\\
%% & ~~$\FF{if0}(\tri; \tri; \tri; \tri)$ & ~ & ~~$\FF{if0}(\iota; \iota; \iota; \iota)$ & ~
%% %& ~~$
%\end{tabular}
%\caption{Syntax from the perspective of fragment providers also includes an internal language and static language.}
%\end{figure*}
%\begin{figure*}
%\small
%\begin{tabular}{r l l}
%external typing context & $\Gamma ::= \emptyset \pipe \Gamma, x \Rightarrow \ttype{tycon}{\hat\sigma}$\\
%\\
% & \textbf{initial} & \textbf{statically normal} \\
%\textbf{static language, cont.} & $\sigma ::= ...$ & $\hat\sigma ::= ...$\\
%argument intro & ~~$\FF{arg}(\Gamma; \hat e)$ & ~~$\FF{arg}(\Gamma; \hat e)$
%\end{tabular}
%\caption{Syntax from the perspective of the semantics / compiler.}
%\end{figure*}
\newcommand{\fkctx}{\mathcal{F}}
\newcommand{\ok}[1]{#1~\mathtt{ok}}
\newcommand{\normalize}[3]{#1 \Downarrow_{#2} #3}
\newcommand{\normalizeX}[2]{\normalize{#1}{\Phi}{#2}}
\newcommand{\serr}[1]{#1~\mathtt{tyerr}}
%\begin{figure*}
%\small
%\begin{tabular}{r l l}
%\textbf{context} & \textbf{syntax} & \textbf{well-formedness}\\
%fragment kind context & $\fkctx ::= \fkctx_0 \pipe \fkctx, \fvar{tycon}[\kappa]~\{\FF{intro}[\kappa]; \FF{elim}[\kappa]\}$ & $\vdash \fkctx$\\
%kinding context & $\Delta ::= \Delta \pipe \Delta, \tvar{x} : \kappa$\\
%fragment context & $\Phi ::= \Phi_0 \pipe \Phi, \fvar{tycon}~\{\hat\delta\}$ & $\vdash \Phi \sim \fkctx$\\
%external typing context & $\Gamma ::= \emptyset \pipe \Gamma, x \Rightarrow \hat\sigma$ & $\vdash_\Phi \Gamma$\\
%internal typing context & $\Omega ::= \emptyset \pipe \Omega, x : \tau$\\
%~
%\end{tabular}
%\\
%\\
%\begin{tabular}{l l l l l}
%\textbf{kinding}& \multicolumn{2}{l}{\textbf{static normalization}} & \multicolumn{2}{l}{\textbf{program compilation}}\\
%$\ok{\rho}$ & $\normalize{\rho}{}{\hat\rho}$ & $\serr{\rho}$ & $\hat\rho \leadsto \iota$\\
%$\Delta \vdash_\fkctx {\phi}\sim\fkctx$ & $\normalizeX{\phi}{\hat\phi}$ & $\serr{\phi}$ & $\hat\phi \sim \Phi$\\
%$\Delta \vdash_\fkctx \delta\sim\{\kappa; \kappa\}$ & $\normalizeX{\delta}{\hat\delta}$ & $\serr{\delta}$ & \multicolumn{2}{l}{\textbf{active typechecking and translation}}\\
%$\Delta \vdash_\fkctx \sigma : \kappa$ & $\normalizeX{\sigma}{\hat \sigma}$ & $\serr{\sigma}$ & $\Gamma \vdash_\Phi \hat e \Leftarrow \hat\sigma \leadsto \iota$ & $\Gamma \vdash_\Phi \hat e \Rightarrow \hat\sigma \leadsto \iota$\\
%$\Delta \vdash_\fkctx \ok{e}$ & $\normalizeX{e}{\hat e}$ & $\serr{e}$ & $\vdash_\Phi \hat\sigma \leadsto \tau$ & $\vdash_\Phi \Gamma \leadsto \Omega$\\
%$\Delta \vdash_\fkctx \ok{\tri}$ & $\normalizeX{\tri}{\iota}$ & $\serr{\tri}$ & \multicolumn{1}{l}{\textbf{internal statics}} & \textbf{internal dynamics}\\
%$\Delta \vdash_\fkctx \ok{\trt}$ & $\normalizeX{\trt}{\tau}$ & $\serr{\trt}$ & $\Omega \vdash \iota : \tau$ & $\iota \mapsto \iota$~~~~~~~~$\iota~\mathtt{val}$
%\end{tabular}
%\caption{Summary of judgements}
%\end{figure*}

%
%\begin{figure*}
%\small
%$
%\begin{array}{lcl}
%\delta_{\texttt{nat}} & := & \quad \FF{iana}~\tlam{opidx}{\kunit+\kunit}{\tlam{tyidx}{\kunit}{\tlam{args}{\klist{\FF{Arg}}}{
% \tsumcase{\tvar{opidx}\\
%&&\quad\quad}{\_}{\tvar{isnil}~\tvar{args}~\titerm{0}\\
%&&\quad\quad}{\_}{\tvar{decons1}~\tvar{args}~\tlam{a}{\FF{Arg}}{\FF{ana}(\tvar{a}; \ttype{nat}{\tunit}; \tvar{x}.\titerm{\iup{\tvar{x}}+1})}}
%}}}\\
%&&\quad \FF{esyn}~\tlam{opidx}{\kunit}{\tlam{tyidx}{\kunit}{\tlam{x}{\kITerm}{\tlam{args}{\klist{\FF{Arg}}}{\tvar{decons2}~\tvar{args}~\tlam{a1}{\FF{Arg}}{\tlam{a2}{\FF{Arg}}{\\
%&&\quad\quad \FF{syn}(\tvar{a1}; \tvar{t1}, \tvar{i1}.\\
%&&\quad\quad\quad \F{let}\tvar{t2} : \kTypeBlur = \ttype{arrow}{(\ttype{nat}{\tunit}, \ttype{arrow}{(\tvar{t1}, \tvar{t1})})}~\F{in} \\
%&&\quad\quad\quad \FF{ana}(\tvar{a2}; \tvar{t2}; \tvar{i2}.(\tvar{t1},\\
%&&\quad\quad\quad\quad \titerm{\iapp{(\ifix{f}{\darrow{\dint}{\dup{\trepof{\tvar{t2}}}}}{\ilam{x}{\dint}{\\&&\quad\quad\quad\quad\quad\quad \FF{if0}(x; \iup{\tvar{i1}}; {
%   %\\&\quad\quad\quad\quad\quad\quad
%   \iapp{\iup{\tvar{i2}}~(x-1)}{(\iapp{f}{(x-1)})}})}}\\
%   &&\quad\quad\quad\quad\quad)}{\iup{\tvar{x}}}})))}}}}}}\\
%&&\quad \FF{rep}~\tlam{tyidx}{\kunit}{\titype{\dint}}\\
%\end{array}
%$
%\caption{\small The definition of the nat type constronstructor.}
%\end{figure*}
%\begin{figure}
%\small
%\begin{flalign}
%& \F{tycon}\fvar{record}~\F{of}\klist{\kpair{\FF{L}}{\kTypeBlur}}~\{\\
%& \quad \FF{iana}~\{\tlam{i}{\kpair{\klist{\kpair{\FF{L}}{\kTypeBlur}}}{\klist{\FF{L}}}}{\tlam{a}{\klist{\Q}}{\\& \quad\quad\tfold{\tvar{zip3}~\tfst{\tvar{i}}~\tsnd{\tvar{i}}~\tvar{a}}
%  {\titerm{\tunit}}{h}{t}{r}{
% \\&\quad\quad\quad\tifeq{\tfst{\tvar{first}~\tvar{h}}}{\tvar{second}~\tvar{h}}{\FF{L}}{
% \\&\quad\quad\quad\quad
% \F{let}~\tvar{x}:\kITerm~=~\FF{ana}(\tvar{third}~\tvar{h}; \tsnd{\tvar{first}~\tvar{h}})
%~\FF{in}
%\\ & \quad\quad\quad\quad \tfold{\tvar{t}}{\tvar{x}
%}{\_}{\_}{\_}{\titerm{(\iup{\tvar{x}},\iup{\tvar{r}})}}  }{\terr}
%%     \F{let}~\tvar{tt}:\FF{TT}=~\FF{syn}(\tsnd{\tsnd{\tvar{h}}}) \F{in}\\
%%\\     & \quad\quad\quad 
%}
%}}\}\\
%& \quad \FF{isyn}~\{\tlam{i}{\klist{\FF{L}}}{\tlam{a}{\klist{\Q}}{
%\\&\quad\quad \tfold{\tvar{zip2}~\tvar{i}~\tvar{a}}{\tden{\titerm{()}}{\ttype{record}{\tnil{{\kpair{\FF{L}}{\kTypeBlur}}}}}}{h}{t}{r}{
%\\&\quad\quad\quad \F{let}~\tvar{htt}:\FF{TT}~=~\FF{syn}(\tsnd{\tvar{h}})
%~\FF{in}
%\\&\quad\quad\quad \F{let}~\tvar{hty}:\kTypeBlur~=~\ttypeof{\tvar{htt}}
%\\&\quad\quad\quad \F{let}~\tvar{ty}:\kTypeBlur~=~\ttype{record}{(\tfst{\tvar{h}}, \tvar{hty})::\tnil{{\kpair{\FF{L}}{\kTypeBlur}}}}
%~\FF{in}
%\\&\quad\quad\quad \tfold{\tvar t}{\tden{\itransof{\tvar{htt}}}{\tvar{ty}}}{\_}{\_}{\_}{
%\\&\quad\quad\quad\quad \tden{\titerm{(\iup{\itransof{\tvar{htt}}},\iup{\itransof{\tvar{r}}})}}{\tvar{ty}}}}
%}}\}\\
%& \quad \FF{esyn}~\{\tlam{i}{\FF{L}}{\tlam{a}{\klist{\Q}}{
% \tvar{decons1}~\tvar{a}~\tlam{ty}{\kTypeBlur}{\tlam{x}{\kITerm}{
%\\&\quad\quad    \tfamcase{\tvar{ty}}{record}{sig}{
%\\&\quad\quad\quad \tfold{\tvar{sig}}{\terr}{h}{t}{r}{
%\\&\quad\quad\quad\quad \tifeq{\tfst{\tvar{h}}}{\tvar{i}}{\FF{L}}{\tfold{\tvar{t}}{\tden{\tvar{x}}{\tsnd{\tvar{h}}}}{\_}{t'}{\_}{
%\\&\quad\quad\quad\quad\quad \tfold{\tvar{t'}}{\tden{\titerm{\ifst{\iup{\tvar{x}}}}}{\tsnd{\tvar{h}}}}{\_}{\_}{\tvar{r'}}{
%\\&\quad\quad\quad\quad\quad\quad \tden{\titerm{\isnd{\iup{{\itransof{\tvar r}}}}}}{\tsnd{\tvar{h}}}}}\\&\quad\quad\quad\quad\hspace{-3px}}{\tvar{r}}
%}\\&\quad\quad\hspace{-3px}}{\terr}
% }}
%}}\\
%& \quad \FF{rep}~\{\tlam{i}{\klist{\kpair{\FF{L}}{\kTypeBlur}}}{   \\& \quad \quad \tfold{\tvar{i}}{\titype{\dunit}}{s}{j}{r}{
%     \tfold{\tvar{j}}{\trepof{\tsnd{\tvar{s}}}}{\_}{\_}{\_}{\\
%   &\quad\quad\quad
%     \titype{\dpair{\dup{\trepof{\tsnd{\tvar{s}}}}}{\dup{\tvar{r}}}}
%     }
%   }}\}\\
%& \}; \\
%& \F{let}\tvar{R} : \kTypeBlur = \ttype{record}{(\ell_1, \ttype{record}{\tnil{{\FF{L}\times\kTypeBlur}}}) :: \tnil{{\FF{L}\times\kTypeBlur}}}~\F{in}\\
%&\F{let}id = \elam{x}{\tvar{R}}{\{\ell_1=~x\cdot\ell_1\} :: \fvar{record}}\\
%&\F{let}triv = id[\{\ell_1 = \{\}\}]\cdot\ell_1
%%&\F{let}one = \FF{intro}[\ell_\text{succ}]
%%&\F{let}plus = \elam{x}{\tvar{nat}}{\elam{y}{\tvar{nat}}{\FF{elim}[()](x; y; \\
%%& \quad \elam{r}{\tvar{nat}}{intro[\ell_\text{succ}](r) : \tvar{nat}}})}
%\end{flalign}
%\caption{The definition of the record type in $\lam@\texttt{lang}$ using the desugarings in Figure \ref{desugaring} and with the addition of simple let bindings for both type and term variables (not shown). Some type-level helper functions also omitted for concision.}
%\label{record-theory}
%\end{figure}

%These correspond to the premises of the \emph{central compilation judgement} $\pkcompiles{\rho}{\iota}$:
%\[
%\inferrule[p-compiles]{
% \emptyset \vdash_{\fvalCtx_0} \rho\\
%%  \progOK{\emptyset}{\fvalCtx_0}{\rho}\\
% \pcompiles{\fvalCtx_0}{\rho}{\iota}
%}{\pkcompiles{\rho}{\iota}}
%\]
%This anchors our exposition; we will describe how it is derived (i.e. how to write a compiler for $\lam@\texttt{lang}$) in the following sections. 
The static semantics are specified by the \emph{bidirectional active typechecking and translation judgements} shown in Fig. \ref{atj}, which relate an expression, $e$, to a {type}, $\sigma$, and a \emph{translation}, $\iota$, under \emph{typing context} $\Gamma$ using imported {fragments} $\Phi$. The judgement form $\atjsynX{e}{\sigma}{\iota}$ specifies \emph{type synthesis} (the type is an ``output''), whereas $\atjanaX{e}{\sigma}{\iota}$ specifies \emph{type analysis} (the type is an ``input''). This can be seen as combining a bidirectional type system (in the style of Pierce and Turner \cite{Pierce:2000:LTI:345099.345100} and a number of subsequent formalisms and languages, e.g. Scala) with an elaboration semantics in the style of the Harper-Stone semantics for Standard ML \cite{Harper00atype-theoretic}. Our language of \emph{internal terms}, $\iota$, includes only functions and lists for simplicity. The form $\iup{\sigma}$ is used as an ``unquote'' operator, and will appear only in intermediate portions of a typing derivation, not in a translation (discussed below).

The first two rows of rules in Fig. \ref{atj} are essentially standard. \rulename{att-subsume} specifies the subsumption principle described in the previous section: if a type can be synthesized, then the term can also be analyzed against that type. We decide type equality purely syntactically here. \rulename{att-var} specifies that variables always synthesize types and elaborate identically. The typing context, $\Gamma$, maps variables to types in essentially the conventional way \cite{pfpl}. The rules $\rulename{att-ana-let}$ and $\rulename{att-syn-let}$ first synthesize a type for the bound value, then add this binding to the context and analyze or synthesize the body of the binding. The translation is to an internal function application, in the conventional manner. $\rulename{att-asc}$ begins by normalizing the provided index and checking that it is a type (Fig. \ref{tleval}, top). If so, it analyzes the ascribed expression against that type. The rules $\rulename{att-ana-slet}$ and $\rulename{att-syn-slet}$ eagerly evaluate the provided static term to a static value, then immediately perform the substitution (demonstrating the phase separation) in the expression before analysis or synthesis proceeds.%The normalization judgement for type-level terms is written $\tau \Downarrow_{\Gamma;\fvalCtx} \tau'$. The kinding rules (not shown here) guarantee that normalization of type-level terms cannot go wrong (we will refine what precisely this means later).

\paragraph{Lambdas}
The rule \rulename{att-ana-lam} performs type analysis on a lambda abstraction, $\lambda(x.e)$. If it succeeds, the translation is the corresponding lambda in the internal language. The type constructor $\textsc{fn}$ is included implicitly in $\Phi$ and must have a type index consisting of a pair of types (pairs are here just lists of length 2 for simplicity). Because both the argument type and return type are known, the body of the lambda is analyzed against the return type after extending the context with the argument type. This is thus the usual type analysis rule for functions in a bidirectional setting.

The rule \rulename{att-syn-idx-lam} covers the case where a lambda abstraction has an incomplete type ascription   providing only the argument type. This corresponds to the concrete syntax seen in the definition of $plus$ in Fig. \ref{atlam-example}. Here, the return type must be synthesized by the body.

These two rules can be compared to the rules in Listing \ref{fn-top}. The main difference is that in @$\lambda$, the language itself manages variables and contexts, rather than the type constructor. This is largely for simplicity, though it does limit us in that we cannot define function type constructors that require alternative or additional contexts. Addressing this in the theory is one avenue for future work.

Function application can be defined directly as a targeted expression, as seen in the example, which we will discuss below.
\paragraph{Fragment Provider Perspective}
\begin{figure}[t]
\small
\centering
$
\begin{array}{l}
\Phi_\text{fn} := \fvar{fn}=\{\FF{analit}=\FF{nil}; \FF{synidxlit}=\FF{nil}; \FF{anatarg}=\FF{nil}; \\
~\quad \FF{syntarg}=\lambda \tvar{tyidx}.\lambda \tvar{ifn}.\lambda \tvar{tmidx}.\lambda \tvar{args}.\\ 
\quad\quad  \tvar{isnil}~\tvar{tmidx}~(\tvar{decons1}~\tvar{args}~\lambda \tvar{arg}. \tvar{decons2} ~\tvar{tyidx}~\lambda \tvar{inty}.\lambda \tvar{rty}.\\
\quad\quad\quad  \FF{ana}(\tvar{arg}; \tvar{inty}; \tvar{ia}.\tvar{pair}~\tvar{rty}~\titerm{\FF{iap}(\iup{\tvar{ifn}};\iup{\tvar{ia}})}))\}
\end{array}
$
\caption{The \textsc{fn} fragment defines function application.}
\label{fn-tycon}
\vspace{10px}
\end{figure}
\begin{figure}[t]
\small
\centering
$
\begin{array}{l}
\Phi_\text{nat} := \fvar{nat}=\{\FF{analit}=\lambda \tvar{tyidx}.\lambda \tvar{tmidx}.\lambda \tvar{args}.\\ 
\quad\quad\tvar{isnil}~\tvar{tyidx}~(\tvar{lbleq}~\tvar{tmidx}~\tvar{lbl}[\texttt{0}]~(\tvar{isnil}~\tvar{args}~ \titerm{\tvar{inil}}));\\
\quad  \FF{synidxlit}=\lambda \tvar{incidx}.\lambda \tvar{tmidx}.\lambda \tvar{args}.\\
\quad\quad \tvar{isnil}~\tvar{incidx}~(\tvar{lbleq}~\tvar{tmidx}~\tvar{lbl}[\texttt{0}]~(\tvar{isnil} ~\tvar{args}~(\tvar{pair}~\FF{nil}~\titerm{\tvar{inil}})));\\
\quad  \FF{anatarg}=\lambda \tvar{tyidx}.\lambda \tvar{i1}.\lambda \tvar{ty}.\lambda \tvar{tmidx}.\lambda \tvar{args}.\\
  \quad\quad \tvar{lbleq}~\tvar{tmidx}~\FF{lbl}[\texttt{rec}]~(\tvar{decons2}~\tvar{args}~
    \lambda \tvar{arg1}.\lambda \tvar{arg2}.\\
    \quad \quad\quad  \FF{ana}(\tvar{arg1}; \tvar{ty}; \tvar{i2}.\FF{ana}(\tvar{arg2}; \tvar{fn2ty}~\FF{ty}[\fvar{nat}](\FF{nil})~\tvar{ty}~\tvar{ty}; \tvar{i3}.\\
    \quad\quad\quad\quad \titerm{\FF{ilistrec}(\iup{\tvar{i1}}; \iup{\tvar{i2}}; x, y.\FF{iap}(\FF{iap}(\iup{\tvar{i3}}; x); y))})));\\
\quad \FF{syntarg} = \lambda \tvar{tyidx}.\lambda \tvar{i1}.\lambda \tvar{tmidx}.\lambda \tvar{args}.\\
\quad\quad \tvar{lbleq}~\tvar{tmidx}~\FF{lbl}[\texttt{s}]~(\tvar{isnil}~\tvar{args}~(\\
\quad\quad\quad \tvar{pair}~\FF{ty}[\fvar{nat}](\tvar{nil})~\titerm{\FF{icons}(\FF{inil};\iup{\tvar{i1}})}))\}
\end{array}
$
\caption{The $\fvar{nat}$ fragment, based on G\"odel's \textbf{T} \cite{pfpl}.}
\vspace{10px}
\label{nat-tycon}
\end{figure}
\begin{figure}[t]
\small
\centering
$
\begin{array}{l}
\Phi_\text{lprod} := \fvar{lprod}=\{\FF{analit} = \lambda \tvar{tyidx}.\lambda \tvar{tmidx}.\lambda \tvar{args}. \\
\quad\quad \FF{listrec}(\tvar{zipexact3}~\tvar{tyidx}~\tvar{tmidx}~\tvar{args}; \titerm{\FF{inil}}; \tvar{h}. \tvar{ri}.\\
\quad\quad\quad \tvar{decons3} ~\tvar{h} ~\lambda \tvar{idxitem}.\lambda \tvar{lbl}.\lambda \tvar{e}.\tvar{decons2}~\tvar{idxitem}~\lambda \tvar{lblidx}. \lambda \tvar{tyidx}. \\
\quad\quad\quad\quad \tvar{lbleq}~\tvar{lbl}~\tvar{lblidx}~(\FF{ana}(\tvar{e}; \tvar{tyidx}; \tvar{i}.\titerm{\FF{icons}(\iup{\tvar{i}}, \iup{\tvar{ri}})})));\\
\quad\FF{synidxlit} = \lambda \tvar{incidx}.\lambda \tvar{tmidx}.\lambda \tvar{args}.\\
\quad\quad \FF{listrec}(\tvar{zipexact2}~\tvar{tmidx}~\tvar{args}; \tvar{pair} ~\FF{nil}~ \titerm{\FF{inil}}; \tvar{h}. \tvar{r}.\\
\quad\quad\quad \tvar{decons2} ~\tvar{h}~\lambda \tvar{lbl}.\lambda \tvar{e}.\tvar{decons2}~\tvar{h}~\lambda \tvar{ridx}.\lambda \tvar{ri}.\FF{syn}(\tvar{e};  \tvar{ty}. \tvar{i}. \\
\quad\quad\quad\quad \tvar{pair}~\FF{cons}(\tvar{pair}~\tvar{lbl}~ \tvar{ty}; \tvar{ridx})~\titerm{\FF{icons}(\iup{\tvar{i}}; \iup{\tvar{ri}})}));\\
\quad \FF{anatarg}=\FF{nil}; \text{\color{gray} (destructuring let could be implemented here)}\\ % 
\quad \FF{syntarg} = \lambda \tvar{tyidx}.\lambda \tvar{i}.\lambda \tvar{lbl}.\lambda \tvar{args}.\\
\quad\quad \tvar{isnil}~\tvar{args}~(\tvar{pair} ~(\tvar{lookup}~\tvar{lbl}~\tvar{tyidx})~\\
\quad\quad\quad \titerm{\FF{iap}(\FF{iap}(nth;\iup{\tvar{i}}); \iup{\tvar{itermofn} ~(\tvar{posof}~\tvar{lbl}~\tvar{tyidx})})})
\end{array}
$
\caption{The $\fvar{lprod}$ fragment (labeled products are like records, but the field order matters; cf. Listing \ref{record}).}
\label{lprod-tycon}
\vspace{10px}
\end{figure}
\begin{figure}[t]
\small
$\fbox{\inferrule{}{\sigma~\mathtt{ty}_\fvalCtx}}$
\vspace{-20px}
\renewcommand{\MathparLineskip}{\lineskiplimit=.6\baselineskip\lineskip=.6\baselineskip plus .2\baselineskip}
\begin{mathpar}
\inferrule[ty]{
  \FF{ty}[\fvar{tycon}](\sigma) \Downarrow_{\emptyctx; \fvalCtx} \FF{ty}[\fvar{tycon}](\sigma)
}{\FF{ty}[\fvar{tycon}](\sigma)~\mathtt{ty}_\fvalCtx}
\end{mathpar}
$\fbox{\inferrule{}{\sigma \Downarrow_{\Gamma;\fvalCtx} \sigma}}$~
%$\fbox{\inferrule{}{\sigma \Downarrow_{\Gamma;\fvalCtx}^= \sigma}}$~
\vspace{-15px}
\begin{mathpar}
%\inferrule[repof]{
% \tau \Downarrow_{\Gamma;\fvalCtx} \ttype{tycon}{\tauidx}\\
% \vdash_\fvalCtx \FF{rep}(\fvar{tycon}) = \taurep\\
% \taurep~\tauidx \Downarrow_{\Gamma;\fvalCtx} \titype{\sigma}
%}{
% \FF{repof}(\tau) \Downarrow_{\Gamma;\fvalCtx} \titype{\sigma}
%}
%
\inferrule[n-ty]{
  \fvar{tycon} \in \text{dom}(\fvalCtx)\\
  \sigma \Downarrow_{\Gamma;\fvalCtx} \sigma'
}{\FF{ty}[\fvar{tycon}](\sigma)~\Downarrow_{\Gamma;\fvalCtx} \FF{ty}[\fvar{tycon}](\sigma')}
\\
\inferrule[n-arg]{ }{\FF{arg}[e]~\Downarrow_{\Gamma;\fvalCtx} \FF{arg}[e]}

\inferrule[n-ana]{
  \sigma_1 \Downarrow_{\Gamma;\fvalCtx} \FF{arg}[e]\\
  \sigma_2 \Downarrow_{\Gamma;\fvalCtx} \sigmat{ty}\\
  \sigmat{ty}~\mathtt{ty}_\fvalCtx\\\\
  \atjanaX{e}{\sigmat{ty}}{\iota}\\
  [\titerm{\iota}/\tvar{x}]\sigma_3 \Downarrow_{\Gamma;\fvalCtx} \sigma_3'
}{
  \FF{ana}(\sigma_1; \sigma_2; \tvar{x}.\sigma_3) \Downarrow_{\Gamma;\fvalCtx} \sigma_3'
}

\inferrule[n-syn]{
    \sigma_1 \Downarrow_{\Gamma;\fvalCtx} \FF{arg}[e]\\
  \atjsynX{e}{\sigma}{\iota}\\\\
    [\sigma/\tvar{x}_{1}, \titerm{\iota}/\tvar{x}_{2}]\sigma_2 \Downarrow_{\Gamma;\fvalCtx} \sigma_2'
}{
  \FF{syn}(\sigma_1; \tvar{x}_{1}.\tvar{x}_{2}.\sigma_2) \Downarrow_{\Gamma;\fvalCtx} \sigma_2'
}

\inferrule[n-quote]{
  \iota \curlyveedownarrow_{\Gamma;\fvalCtx} \iota'
}{\titerm{\iota} \Downarrow_{\Gamma;\fvalCtx} \titerm{\iota'}}
\end{mathpar}
$\fbox{\inferrule{}{\iota \curlyveedownarrow_{\Gamma;\fvalCtx} \iota}}$
\vspace{-15px}
\begin{mathpar}

\inferrule[q-uq]{\sigma \Downarrow_{\Gamma;\fvalCtx} \titerm{\iota}}{\iup{\sigma} \curlyveedownarrow_{\Gamma;\fvalCtx} \iota}

\inferrule[q-x]{ }{x \curlyveedownarrow_{\Gamma;\fvalCtx} x}

\inferrule[q-lam]{\iota \curlyveedownarrow_{\Gamma;\fvalCtx} \iota'}
{\lambda(x.\iota) \curlyveedownarrow_{\Gamma;\fvalCtx} \lambda (x.\iota')}
%
%\inferrule[q-ap]{\iota_1 \curlyveedownarrow_{\Gamma;\fvalCtx} \iota_1'\\
%\iota_2 \curlyveedownarrow_{\Gamma;\fvalCtx} \iota_2'}{
%\FF{iap}(\iota_1; \iota_2) \curlyveedownarrow_{\Gamma;\fvalCtx} \FF{iap}(\iota_1'; \iota_2')}
\end{mathpar}
\caption{Selected normalization rules for the static language.}
\vspace{10px}
\label{tleval}
\end{figure}

Fragment providers define type constructors by defining ``methods'', $\delta$, that control analysis and synthesis literals and targeted expressions. These are static functions invoked by the final four rules in Fig. \ref{atj}, which we will describe next. The type constructors $\fvar{fn}$, $\fvar{nat}$ and $\fvar{lprod}$ are shown in Figs. \ref{fn-tycon}-\ref{lprod-tycon}, respectively. They use helper functions for working with labels ($\tvar{lbleq}$), lists ($\tvar{isnil}$, $\tvar{decons1}$, $\tvar{decons2}$, $\tvar{decons3}$, $\tvar{zipexact2}$, $\tvar{zipexact3}$, and $\tvar{pair}$), lists of pairs interpreted as finite mappings ($\tvar{lookup}$ and $\tvar{posof}$) and translating a static representation of a number to an internal representation of that number ($\tvar{itermofn}$). We also assume an internal function $nth$ that retrieves the $n$th element of a list. All of these are standard or straightforward and omitted for concision. Failure cases for the static helper functions evaluate to $\FF{fail}$. Derivation of the typing judgement does not continue if a $\FF{fail}$ occurs (corresponding to an \lstinlinep{atlang.TypeError} propagated directly to the compiler).

\subparagraph{Literals} The rule $\rulename{att-ana-lit}$, invokes the $\FF{analit}$ method of the type constructor of the type the literal is being analyzed against, asking it to return a translation (a value of the form $\titerm{\iota}$). The type and term index are provided, as well as a list of \emph{reified arguments}: static values of the form $\FF{arg}[e]$. The $\fvar{fn}$ type constructor does not implement this (functions are introduced only via lambdas). The $\fvar{nat}$ type constructor implements this by checking that the term index was the label corresponding to $\texttt{0}$ and no arguments were provided. The $\fvar{lprod}$ type constructor is more interesting: it folds over each corresponding item in the type index (a pair consisting of a label and a type), the term index (a label) and the argument list, checking that the labels match and programmatically analyzing the argument against the type it should have. A reified argument $\sigma_1$ against a type $\sigma_2$, binding the translation to $\tvar{x}$ in $\sigma_3$ if successful (and failing otherwise) with the static term $\FF{ana(\sigma_1; \sigma_2; \tvar{x}.\sigma_3)}$, the semantics of which are in Fig. \ref{tleval}. Labeled products translate to lists by recursively composing the translations of the field values using the ``unquote'' form, $\iup{\sigma}$, which is eliminated during normalization (also seen in Fig. \ref{tleval}). This function can be compared to the method \lstinlinep{ana_Dict} in Listing \ref{record}.

Literals with an incomplete type ascription can synthesize a type by rule \rulename{att-syn-idx-lit}. The $\FF{synidxlit}$ method of type constructor of the partial ascription is called with the incomplete type index, the term index and the arguments as above, and must return a pair consisting of the complete type index and the translation. Again, $\fvar{fn}$ does not implement this. The $\fvar{nat}$ type constructor supports it, though it is not particularly interesting, as $\fvar{nat}$ is always indexed trivially, so it follows essentially the same logic as in $\FF{analit}$. The $\fvar{lprod}$ type constructor is more interesting: in this case, when the incomplete type index is trivial (as in the example in Fig. \ref{atlam-example}), the list of pairs of labels and types must be synthesized from the literal itself. This is done by programatically synthesizing a type and translation for a reified argument using $\FF{syn(\sigma_1; \tvar{x}.\tvar{y}.\sigma_2)}$, also specified in Fig. \ref{tleval}. The type index and translation are recursively formed. This can be compared to the class method \lstinlinep{syn_idx_Dict} in Listing \ref{record}.

\subparagraph{Targeted Terms} Targeted terms are written $\FF{targ}[\sigmat{tmidx}](e_\text{targ}; \overline{e})$. The type constructor of the type recursively synthesized by the \emph{target}, $e_\text{targ}$, is delegated control over analysis and synthesis via the methods $\FF{anatarg}$ and $\FF{syntarg}$, respectively, as seen in rules \rulename{att-ana-targ} and \rulename{att-syn-targ}. Both receive the type index, the translation of the target, the term index and the reified arguments. The former also receives the type being analyzed and only needs to produce a translation. The latter must produce a pair consisting of a type and a translation.

The $\fvar{fn}$ type constructor defines function application by straightforwardly implementing $\FF{syntarg}$.  Because of subsumption, $\FF{anatarg}$ need not be separately defined.

The $\fvar{nat}$ type constructor defines the successor operation synthetically and the recursor operation analytically (because it has two branches that must have the same type). The latter analyzes the second argument against a function type, avoiding the need to handle binding itself. Natural numbers translate to lists, so the $\FF{nat}$ recursor can be implemented straightforwardly using the list recursor. In a practical implementation, we might translate natural numbers to integers and use a fixpoint computation instead.

The $\fvar{lprod}$ type constructor defines the projection operation synthetically, using the helper functions mentioned above to lookup the appropriate item in the type index. Note that one might also define an analytic targeted operation on labeled products corresponding to pattern matching, e.g. $\FF{let}~\{\texttt{a}{=}x, \texttt{b}{=}y\}=r~\FF{in}~e$. @$\texttt{lang}$ supports this using Python's syntax for destructuring assignment, but we must omit the details.

\paragraph{Metatheory} The formulation shown here guarantees that type synthesis actually produces a type, given well-formed contexts. The definitions are straightforward and the proof is a simple induction with simple lemmas about the normalization semantics. %We plan to prove the full proofs in a companion tech report (available upon request).

\begin{theorem}[Synthesis]
If $\fvalCtx~\mathtt{fragment}$ and $\Gamma~\mathtt{ctx}_\fvalCtx$ and $\atjsynX{e}{\sigma}{\iota}$ then $\sigma~\mathtt{ty}_\fvalCtx$.
\end{theorem}

We can also prove that importing additional type constructors cannot change the semantics of a previously well-typed term, assuming that naming conflicts have been resolved by some extrinsic mechanism. Indeed, the proof is an essentially trivial induction because of the way we have structured our mechanism. The type constructor delegated responsibility over a term is deterministically determined irrespective of the structure of $\fvalCtx$. We plan to provide full proofs of these theorems in a technical report.

\begin{theorem}[Unambiguous Extension]
If $\fvalCtx~\mathtt{fragment}$ and $\Gamma~\mathtt{ctx}_\fvalCtx$ and $\atjsynX{e}{\sigma}{\iota}$ and $\fvalCtx'~\mathtt{fragment}$ and $\text{dom}(\fvalCtx) \cap \text{dom}(\fvalCtx') = \emptyset$ then ${\Gamma}\vdash_{\fvalCtx;\fvalCtx'}{e}\Rightarrow{\sigma}\leadsto{\iota}$.
\end{theorem}
%Metatheoretic guarantees about the translation cannot be provided in the formulation as given. However, inserting straightforward checks to guarantee that the translation is a closed term, and provide simple mechanisms for hygiene, would be simple, but are omitted.
%\noindent
%\subsubsection{Rules} We briefly describe each rule and relevant additional judgements below:
%
%\noindent
%The rule $\rulename{att-lam}$ says that lambda terms can only be analyzed against arrow types and translate to lambda terms in the internal language.
%
%The rule $\textsc{\textbf{att-intro-ana}}$ says that generalized introductory forms can only be analyzed against a type. That type constructor is consulted to extract its \emph{introductory operator definition}, written (e.g. in Figure \ref{nat-atfrag}) after the $\FF{iana}$ keyword as a type-level function. This function is given the provided operator index, the type's type index and a list encapsulating the operator's  \emph{arguments}. An argument is reified as a type-level term of the form $(\Gamma, e)?$ and has kind $\FF{Arg}$. Note that only the compiler should construct such a term (in practice, this could be enforced purely syntactically, so we do not enforce this judgmentally here). The operator definition is responsible for producing a translation, or evaluating to $\terr$ if this is not possible. A translation is simply a quoted internal term, written $\titerm{\iota}$, with kind $\kITerm$. Like quoted internal types, quoted internal terms support an unquote form, written $\iup{\tau}$, which is normalized away (in a capture-avoiding manner, not shown).
%
%Natural numbers have two introductory forms, each indexed trivially. Because there is only one generalized introductory form, we instead index the operator by a simple sum kind with two trivial cases, $\kunit + \kunit$. The first case corresponds to the zero operator and the second to the successor. In the introductory operator definition, we see in the first case that the translation $0$ is produced after checking that no arguments were provided (using a simple helper function, $\tvar{isnil}$, not shown). In the second case, we first extract the single argument (using the helper function $\tvar{decons1}$, not shown). We then analyze it against the type $\ttype{nat}{\tunit}$ using one of the elimination forms for arguments, $\FF{ana}(\tau_1; \tau_2; \tvar{t}_{trans}.\tau_3)$. If it succeeds, we construct the appropriate translation based on the translation of the argument. The normalization rule for successful analysis, $\textsc{ana}$, is shown in Figure \ref{tleval}. 
%
%Note also that lambda is technically the introductory form for the $\fvar{arrow}$ type constructor, but because it requires manipulating the context, it must be built in to @$\lambda$. We do not currently support type system fragments that require adding or manipulating existing contexts.
%
%The rule $\rulename{att-elim-syn}$ says that elimination forms synthesize a type. This is done again by consulting a type-level function, called the \emph{eliminatory operator definition}, associated with a type constructor, here the type constructor of the type  recursively synthesized for the primary operand, $e$. This definition is invoked with the type index, the operator index, the translation of the primary operand and a list of arguments. Because elimination forms synthesize a type, this definition must produce both a type and a translation. Representational consistency is then checked.
%
%Note that the elimination form for the $\fvar{arrow}$ type constructor can be defined in this manner (because it does not require binding variables), so application is simply syntactic sugar, rather than a built-in operator in the external language.
%
%The constructor context, $\Phi$, tracks user-defined type  constructors introduced in the ``imported'' fragments. %Note that fragments can also export type-level values bound to type-level variables (written in bold font, e.g. $\tvar{nat}$, $\tvar{s}$, $\tvar{z}$, etc.), but substitution will be performed prior to typechecking, so they need not be tracked by $\Phi$. % Weakening and exchange of the constructor context closely related to the issue of conservativity that we will return to. Each constructor is identified by name, so there is no analog to contraction. 
% This form of semantics can be seen as lifting into the language specification the first stage of a type-directed compiler like the TIL compiler for Standard ML \cite{tarditi+:til-OLD} and has some parallels to the Harper-Stone semantics for Standard ML \cite{Harper00atype-theoretic}. There, as in Wyvern, external terms were given meaning by elaboration from the EL to an IL having the same type system. Here the two languages have different type systems, so we call it a translation rather than an elaboration (arranging the judgement form slightly differently to emphasize this distinction, cf. above).

%In @$\lambda$, the internal language (IL) provides partial functions (via the generic fixpoint operator of Plotkin's PCF), simple product  types and integers for the sake of our example (and as a nod toward practicality on contemporary machines). In practice, the internal language could be any typed  language with a specification for which type safety and decidability of typechecking have been satisfyingly determined. In Sec. \ref{ace}, we will see how the internal language can itself be made user-definable. The internal type system serves as a ``floor'': guarantees that must hold for terms of any type (e.g. that out-of-bounds access to memory never occurs) must be maintained by the internal type system. User-defined constructors can enforce invariants stronger than those the internal type system maintains at particular types, however. Performance is also ultimately limited by the internal language and downstream compilation stages that we do not here consider (safe compiler extension has been discussed in previous work, e.g. \cite{conf/pldi/TatlockL10}).
%
%The external language has a fixed syntax with six forms: variables, $\FF{let}$-bindings, lambda terms, type ascription and generalized introductory and elimination forms. As we will see, the generalized introductory form is given meaning by the type it is analyzed against (similar to the protocol for TSLs in Wyvern), and the elimination form is given meaning by the type of the external term being eliminated ($e$). This represents an internalization into the language of Gentzen's inversion principle \cite{gentzen}\todo{what to cite for this?}. Fig. \ref{desugaring} shows how to recover more conventional introductory and elimination forms by a purely syntactic desugaring.%\footnote{Note that for even more flexibility, we could also include TSLs, but we choose to avoid that for simplicity (and to show that, strictly speaking, one need not have an extensible syntax to have an extensible type system, and \textit{vice versa}).}
%
%\subsection{Types and Type-Level Computation}\label{types}
%@$\lambda$ supports, and makes extensive use of, simply-kinded type-level computation. Specifically, type-level terms, $\tau$, themselves form a typed lambda calculus. The classifiers of type-level terms are called \emph{kinds}, $\kappa$, to distinguish them from  \emph{types}, which are  type-level values of kind $\kTypeBlur$. %In @\texttt{lang}, the type-level language (together with the level of programs) is written in Python, which can be thought of as having a rather limited kind system (with one kind, \lstinlinep{dyn}). Here, we are able to more precisely discuss the kinds of values in the type-level language. 
%As in Sec. \ref{att}, types are formed by applying a \emph{type constructor} to an \emph{index}. User-defined type constructors are declared in fragment definitions using $\FF{tycon}$. Each constructor in the program must have a unique name, written e.g. \fvar{nat} or \fvar{lprod}. %\footnote{We assume naming conflicts can be avoided by some extrinsic mechanism.} 
%A type constructor must also declare an \emph{index kind}, $\kappat{tyidx}$. A type is introduced by applying a type constructor to an index of this kind, written $\ttype{Tycon}{\taut{tyidx}}$.  For example, the type of natural numbers is indexed trivially (i.e. by kind $\kunit$), so it is written $\ttype{Nat}{\tunit}$.
%
% To permit the embedding of interesting type systems, the type-level language includes several kinds other than $\kTypeBlur$. We lift several functional data structures to the type level: here, only unit ($\kunit$), binary products ($\kpair{\kappa_1}{\kappa_2}$), binary sums ($\ksum{\kappa_1}{\kappa_2}$) and lists ($\klist{\kappa}$), in addition to labels (introduced as $\ell$, possibly with a subscript, having kind \FF{L}).  The type constructor $\fvar{nat}$ is indexed trivially because there is only one natural number type, but $\fvar{lprod}$ would be indexed by a list of pairs of {labels}  and types. The type constructor $\fvar{arrow}$ is included in the initial constructor context, $\fvalCtx_0$, and has index kind $\kpair{\kTypeBlur}{\kTypeBlur}$.
% As with the internal language, in practice, one could include a richer programming language and retain the spirit of the calculus, as long as it does not introduce general recursion at the type level. For example, our desugarings add support for number literals and string literals, which require adding numbers and strings to the type-level language (and providing a means for lifting them from the type-level language to the internal language, as we will discuss).
%  
% %We see a record type, abbreviated $\tvar{R}$, constructed on line 26. 
%
%The kind $\kTypeBlur$ also has an elimination form, $\tfamcase{\tau}{Tycon}{x}{\tau_1}{\tau_2}$ allowing the extraction of a type index by case analysis against a contextually-available type constructor. To a first approximation, one might think of type constructors as constructors of a built-in open datatype \cite{conf/ppdp/LohH06}, $\kTypeBlur$, at the type-level. Like open datatypes, there is no notion of exhaustiveness so the default case is required for totality. %We will see where this is used shortly.
%
%
%
%%We will write the kind of types as $\kTypeBlur$, though it is also written $\star$ or \lstinlinep{Type} in various similarly structured languages (see Sec. \ref{related-work}). 
%% Rather than there being a fixed set of type constructors, we allow the programmer to declare new type  constructors, and give the static and dynamic semantics of their associated operators, by writing type-level functions. In the semantics for this calculus, our kind system combined with techniques borrowed from the typed compilation literature and a form of type abstraction allow us to prove strong type safety, decidability and conservativity theorems.
%
%
%
%% and integers ($\dint$). We also include labels ($\klabel$), written in a slanted font, e.g. $\tlabel{myLabel}$, which are string-like values that only support comparison and play a distinguished role in the expanded syntax, as we will later discuss. Our first example, $\fvar{nat}$, is indexed trivially, i.e. by unit kind, $\kunit$, so there is only one natural number type, $\ttype{nat}{\tunit}$, but we will show examples of type constructors that are indexed in more interesting ways in later portions of this work. For example, $\fvar{LabeledTuple}$ has index kind $\klist{\kpair{\klabel}{\kTypeBlur}}$. 
% 
% Type constructors are not first-class; they do not themselves have arrow kind as in some kind systems (e.g.  \cite{watkins2008specifying}; Ch. 22 of \emph{PFPL} describes a related system \cite{pfpl}). The type-level language does, however, include total functions of arrow kind, written $\karrow{\kappa_1}{\kappa_2}$. Type constructor application can be wrapped in a type-level function to emulate a first-class or uncurried version of a type constructor for convenience.% (indeed, such a wrapper could be generated automatically, though we do not do so). 
%
%Two type-level terms of kind $\kTypeBlur$ are equivalent if they apply the same constructor, identified by name, to equivalent indices. Going further, we ensure that deciding type equivalence requires only checking for syntactic equality after normalization by imposing the restriction that equivalence at a type constructor's index kind must be decidable in this way. Our treatment of equivalence in the type-level language is thus quite similar to the treatment of term-level equality using ``equality types'' in a language like Standard ML.
%% A kind $\kappa$ is an  \emph{equality kind} if $\kEq{\kappa}$ can be derived (see appendix). 
%Conditional branching on the basis of equality at an equality kind can be performed in the type-level language. Equivalence at arrow kind is not decidable by our criteria, so type-level functions cannot appear within type indices. This also prevents general recursion from arising at the type level. Without this restriction, a type-level function taking a type as an argument could ``smuggle in'' a self reference as a type index, extracting it via case analysis (continuing our analogy to open datatypes, this is closely related to the positivity condition for inductive datatypes in total functional languages like Coq).% as maintaining the metatheoretic guarantee that typing respects type equivalence would impose a substantial burden in such a setting.% (a na\"ive approach to this would impose non-trivial extrinsic proof obligations onto extension developers that, unlike in others in this thesis, could threaten type safety).

%Every type constructor also defines type-level functions called $\FF{iana}$ and $\FF{esyn}$, which we will describe below, and a \emph{representation schema}, a type-level function that associates with every type an internal type. We will return to this after introducing operators.
%
%\subsection{Core External Forms and Desugaring}\label{opcons}
%The syntax for external terms (Figure \ref{grammar}) contains variables, $\lambda$ terms, three generalized introductory forms and a single generalized elimination form. The introductory forms are either unascribed, ascribed with a type or ascribed with a type constructor, as in our discussion of @\texttt{lang}. These generalized forms take a single a type-level value as an {index} and $n \geq 0$ arguments, which are other external terms. To better motivate this choice, we can give a purely syntactic desugaring of a Python-like syntax with labels to these forms, shown in Figure \ref{desugaring}. It is instructive to rewrite lines 27-28 of Figure \ref{record-theory} using these desugarings.% In @\texttt{lang}, desugarings can be user-defined 
%
%User-defined operator constructors are declared using \textsf{opcon}.  For reasons that we will discuss, our calculus associates every operator  constructor with a type constructor. The \emph{fully-qualified name} of every operator constructor, e.g. $\fvar{nat}.\opvar{z}$, must be unique. Operator constructors, like type constructors, declare an index kind, $\kappaidx$. In our first example, all the operator constructors are indexed trivially (by index kind $\kunit$), but other examples use more interesting indices. For example, in SML, the projection operator \lstinlinep{#3} can be applied to an $n$-tuple, $e$, iff $n \geq 3$. Note that it thus cannot be a function with a standard arrow type. Notionally, \lstinlinep{#} is an operator constructor and \lstinlinep{3} is its index. In an active embedding of $n$-tuples into $\lam@\texttt{lang}$, this would be written $\eop{Tuple}{prj}{3}{e}$ (we will nearly recover ML's syntax later). $\fvar{LabeledTuple}.\opvar{prj}$ is the operator constructor used to access a field of a labeled tuple, so it has index kind $\klabel$. An operator itself is, notionally, selected by indexing an operator constructor, e.g. $\fvar{nat}.\opvar{s}\langle \tunit \rangle$, but technically neither operator constructors nor operators are first-class at any level (additional machinery would be needed, e.g. an \textsf{Op} kind, but this is not fundamental to our calculus). 
%Instead, in the external language, an operator constructor is applied by simultaneously providing an index and  $n \geq 0$ \emph{arguments}, written $\eop{Tycon}{op}{\tauidx}{\splat{e}{1}{n}}$\footnote{It may be helpful to distinguish between type/operator constructors and \emph{term formers}. There are term formers at all levels in the calculus. For example, operator constructor application and $\lambda$ are  external term formers, and type constructor application is a type-level term former. We might write these following Harper's conventions for abstract syntax to highlight this distinction \cite{pfpl}: $\mathtt{lam}[\tau](x.e)$, $\mathtt{ocapp}[\fvar{Tycon}, \opvar{op}, \tauidx](\splat{e}{1}{n})$ and $\texttt{tcapp}[\fvar{Tycon}](\tau)$.}. For example, on line 18 of Fig. \ref{nat}, we see the operator constructors $\fvar{nat}.\opvar{z}$ and $\fvar{nat}.\opvar{s}$ being applied to compute $two$. %\footnote{Although our focus here is entirely on semantics, a brief note on syntax: in the expanded syntax, the trivial indices and empty argument lists can be omitted, so we could write \texttt{Nat.s(Nat.s(Nat.z))}. With the ability to ``open'' a type's operators into the context, we could shorten this still to \texttt{s(s(z))}. Alternatively, with the ability to define a TSL in a manner similar to that in Sec. \ref{aparsing}, we might instead just write \texttt{2}.}

%The expressive and metatheoretic power of the calculus arises from how the rules for the active typing judgement handle these  generalized forms. Rather than fixing the specification of a finite collection of operator constructors and tasking the \emph{compiler} with deciding a typing derivation on its basis, the specification instead delegates to a type-level function associated with a type constructor. In the core calculus, this is one of three functions, called \FF{iana}, \FF{isyn} and \FF{esyn} in the grammar.\footnote{This means that any particular type constructor only supports a single introduction and elimination desugaring. This is a minor inconvenience in some cases that is resolved in @\texttt{lang} by the use of methods.} We see how this is done with the rules for the active typing judgements, given in Figure \ref{atj}.

%The rule $\rulename{att-lam}$ must take into account the fact that, because we support type-level computation, the type annotation on the argument may not be in normal form. Thus, we evaluate it to normal form. Note that the kinding rules (not shown) will guarantee that, because $\tau_1$ is of kind $\kTypeBlur$, its normal form, $\tau'_1$, is of the form $\ttype{tycon}{\tauidx}$ for some normal $\tauidx$. To generate an appropriate internal type annotation in the translation, we need to compute the representation type of $\tau_1'$. This involves calling the representation function associated with its type constructor (rule \fvar{repof} in Figure \ref{tleval}). The form $\titerm{\sigma}$ is a \emph{quoted internal type}. Note in the syntax that there is an unquote form, $\dup{\tau}$, which allows us to compose internal types compositionally without needing to expose an elimination form. Indeed, it is an interesting facet of our calculus that we never need to examine syntax trees directly to implement extensions. The evaluation semantics remove quotations, so the normal form of a quoted internal type contains an internal type with no quotations.
%
%Lambda functions are the only introductory form requiring special support in the calculus (because they need to  manipulate the context; see Discussion). The next three rules show how any other abstractions that we define (e.g. records, decimals, etc.) make use of a generalized introductory form. 
%
%The rule  $\rulename{att-i-unasc}$ shows that unascribed introductory forms can only be analyzed against a type. Given such a type, the rule extracts a definition, named $\FF{iana}$, from the type constructor (cf. the \lstinlinep{ana_Dict} and \lstinlinep{trans_Dict} methods earlier). It calls this function with a pair containing the operator and type index and a list of \emph{reified arguments}, which have kind $\Q$ and introductory form $(\Gamma; e)?$. These are only constructed by the compiler (there would be no corresponding form in the concrete syntax). Their purpose is to allow the definition to programatically invoke synthesis and analysis as needed using the $\FF{ana}$ and $\FF{syn}$ operators. We can see the $\fvar{record}$ type constructor doing so on line 5 of Figure \ref{record-theory} to ensure that the field value provided as an argument has the same type as the corresponding label (accomplished by simultaneously folding over all three pieces of input data). The rule for performing analysis, $\fvar{ana}$, is in Figure \ref{tleval}. If it succeeds, analysis returns a \emph{translation}, which is a quoted internal term with kind $\kITerm$ and introductory form $\titerm{\iota}$. Like quoted internal types, there is an unquote form that is eliminated during evaluation. The operator definition uses this to construct a translation for the record. As before, empty records translate to units and records with a single field are unadorned. In this example, records with two or more fields translate to nested tuples. 
%
%Because we have a representation type associated with the type, we can check that the translation is \emph{representationally consistent} (the final premise). As we will discuss, representational consistency combined with type safety of the internal language implies type safety overall (it arises as a strengthening of the inductive hypothesis needed to prove type safety, and is closely related to work on \emph{type-preserving compilation} in the TIL compiler for Standard ML, \cite{tarditi+:til-OLD}).
%
%The next rule, $\rulename{att-i-asc-ty}$ states that introductory forms ascribed with a type are analyzed against that type.
%
%Introductory forms ascribed with a type constructor can, according to the final introductory rule, $\rulename{att-i-asc-tycon}$, synthesize a type via the definition $\FF{isyn}$. It is passed the operator index (there is no type index, since we only have a type constructor) and a list of arguments, as before. In this case, it must return not just a translation, but also a type. We use the form $\tden{\tau_2}{\tau_1}$ for such a pairing, which has kind $\FF{TT}$. The type and translation can be extracted from it using the appropriately named elimination forms (indeed, as given, it is merely a pair, but we give it special syntax for clarity -- it looks like the conclusion of the typing judgement -- and for other reasons that will become clear in future work). To synthesize a type from a list of labels and arguments, we must be able to synthesize types from the arguments. The $\FF{syn}(\tau)$ operator permits this, per rule $\fvar{syn}$, seen being used in Listing \ref{record-theory}. We check representational consistency of the result, as before.
%
%Finally, we show the rule for elimination forms. It operates by first synthesizing a type for the ``primary'' subterm, then extracting the $\FF{esyn}$ definition from its type constructor. As before, it is called with the arguments and representational consistency is checked. Note that the primary subterm is itself the first argument (though we have already synthesized a type for it, it is more uniform and clear to allow the operator  to do so again, which is done in our example by the $\tvar{decons1}$ helper function).

%\subsection{Representational Consistency Implies Type Safety}\label{repcon}
%A key invariant that our operator definitions maintain is that well-typed external terms of type $\ttype{Nat}{\tunit}$ always translate to well-typed internal terms of internal type $\dint$, the representation type of $\ttype{Nat}{\tunit}$. Verifying this for the zero case is simple. For the translation produced by the successor case to be of internal type $\dint$ requires that $\iup{\tvar{x}}$ be of internal type $\dint$. Because it is the result of analyzing an argument against $\ttype{Nat}{\tunit}$, and the only other introductory form is the zero case, this holds inductively. We will discuss the recursor later, but it also maintains this invariant inductively.
%
%This invariant would not hold if, for example, we allowed the type and translation: $$({{\ttype{Nat}{\tunit},\titerm{(0, ())}}})$$
%\noindent
%In this case, there would be two different internal types, $\dint$ and $\dpair{\dint}{\dunit}$,  associated with a single external type, $\ttype{Nat}{\tunit}$. This would make it impossible to reason \emph{compositionally} about the translation of an external term of type $\ttype{Nat}{\tunit}$, so our implementation of the successor would produce ill-typed translations in some cases but not others. Similarly, we wouldn't be able to write functions over all natural numbers because there would not be a well-typed translation to give to such a function. This violates type safety: there are now well-typed external terms, according to the active typechecking and translation judgement, for which evaluating the corresponding translation would ``go wrong''. 
%
%To reason compositionally about the semantics of well-typed external terms when they are given meaning by translation to a typed internal language, the system must maintain the following property: for every  type, $\tau$, there must exist an internal type, $\sigma$, called its \emph{representation type}, such that the translation of every external term of type $\tau$ has internal type $\sigma$. This principle of \emph{representational consistency} arises essentially as a strengthening of the inductive hypothesis necessary to prove that all well-typed external terms translate to well-typed internal terms because $\lambda$ and successor are defined compositionally. It is closely related to the concept of \emph{type-preserving compilation} developed by Morrisett et al. for the TIL compiler for SML \cite{tarditi+:til-OLD}, here lifted into the language. Our judgements check this extension correctness property directly by typechecking each translation produced by a user extension (the translations of, for example, lambda terms will be inductively representationally consistent, so no additional check is needed).%That is, instead of being used as a necessary condition for compiler correctness, we are using it as a sufficient condition for type safety. 
%%We emphasize the distinction between translation (which in our calculus endows terms with a dynamic semantics) and compilation (which must preserve the dynamic semantics of terms). %In the presence of extensions, checking for representational consistency becomes subtle if we wish to guarantee conservativity, as we will discuss in the next subsection.
%
%%For the semantics to ensure that representational consistency is maintained by all operator definitions, we require, as briefly mentioned in Sec. \ref{types}, that each type constructor declare a \emph{representation schema} after the keyword \textsf{schema}. This is a type-level function of kind $\karrow{\kappaidx}{\kIType}$, where $\kappaidx$ is the index kind of the type constructor. The kind $\kIType$ has introductory form $\titype{\bar \sigma}$ and no elimination form, and is similar to the kind $\kITerm$ introduced above. There are two forms of \emph{translational internal type}, $\bar \sigma$, that do not correspond to forms in $\sigma$. As with translational internal terms, these are included to allow an internal type to be formed compositionally:
%%\begin{enumerate}
%%\item $\dup{\tau}$ splices in another translational internal type $\tau$
%%\item $\trepof{\tau}$ refers to the representation type of type $\tau$
%%\end{enumerate}
%%
%%When the semantics (i.e. the compiler) needs to determine the representation type of the type $\ttype{Tycon}{\tauidx}$ it simply applies the representation schema of $\fvar{Tycon}$ to ${\tauidx}$. Note that the representation type of a type cannot be extracted directly from within the type-level language, again for reasons that we will discuss below. 
%%
%%These additional forms are not needed by the representation schema of $\fvar{nat}$  because it is trivially indexed. In Fig. \ref{tuple}, we will discuss an example of the type constructor $\fvar{Tuple}$, implementing the semantics of $n$-tuples by translation to nested binary products. Here, the representation schema refers to the representations of the tuple's constituent types and is computed by folding over a list, so both of these are used\todo{add this example from appendix of ESOP}. 
%
%Translational internal terms can contain translational internal types. We see this in the definition of the recursor on natural numbers. The type assignment is the arbitrary type, $\tvar{t2}$. We cannot know what the representation type of $\tvar{t2}$ is, so we refer to it abstractly using $\trepof{\tvar{t2}}$. Luckily, the proof of representational consistency of this operator is parametric over the representation type of $\tvar{t2}$.

%\subsection{Metatheory}\label{metatheory}
%\newcommand{\F}[1]{\textsf{#1}~}
%\newcommand{\FF}[1]{\textsf{#1}}
%\newcommand{\Q}{\FF{Arg}}
%\renewcommand{\tnil}[1]{[]}
%\begin{figure}[t]
%\small
%$$\begin{array}{rcl} 
%%\textbf{programs} & 
%\rho & ::= & \F{tycon}\fvar{tycon}~\F{of}\kappaidx~{\{}TC{\}}; \rho \pipe e\\ %\pfam{\familyDf}{\progsort}
%%\\&  \pipe & \pdef{t}{\kappa}{\tau}{\progsort}  \pipe e
%%\pipe \pdef{t}{\kappa}{\tau}{\progsort} 
% TC & ::= & \F{iana} \{\tau\}; \F{isyn} \{\tau\}; \F{esyn} \{\tau\}; \F{rep} \{\tau\}\\
%\\ 
%e & ::= & \evar{x} \pipe \elam{\evar{x}}{\tau}{e} \pipe I \pipe I : \tau \pipe I :: \fvar{tycon} \pipe E\\
%I & ::= &  \FF{intro}[\tauidx](\splat{e}{1}{n})\\
%E & ::= & e\cdot\FF{elim}[\tauidx](\splat{e}{1}{n})\\
%\\
% \tau  & ::=   &   \tvar{t} \pipe 
%                           \tlam{t}{\kappa}{\tau} \pipe 
%                           \tapp{\tau_1}{\tau_2}  
%\\&\pipe&                      
%                           \tnil{\kappa} \pipe \tcons{\tau_1}{\tau_2} \pipe 
%                                      \tfold{\tau_1}{\tau_2}{h}{t}{r}{\tau_3}
%                           \\
%     & \pipe &     \ell \pipe 
%   \tunit \pipe 
%                           \tpair{\tau_{1}}{\tau_{2}} \pipe 
%                           \tfst{\tau} \pipe 
%                           \tsnd{\tau} \pipe \cdots 
%                           \\  
%%     & \pipe & \tinl{\kappa_2}{\tau_1} \pipe \tinr{\kappa_1}{\tau_2} \\
%%     &\pipe& \tsumcase{\tau}{t}{\tau_1}{t}{\tau_2} \\
%&\pipe &   \ttype{tycon}{\tauidx} \\&\pipe& \tfamcase{\tau}{tycon}{x}{\tau_1}{\tau_2}\\
%&\pipe &           \tifeq{\tau_{1}}{\tau_{2}}{\kappa}{\tau_{3}}{\tau_{4}} 
%%                            \\                          %       & & \pipe & \tfamcase{\tau}{Fam}{x}{\tau_1}{\tau_2}\\
%%                                                
%\\      &  \pipe & \tden{\tau_2}{\tau_1} \pipe \terr %\pipe \tden{\ibar}{\tau}^{\checkmark} 
%\pipe \ttypeof{\tau} \pipe \itransof{\tau} \\% \tdencase{\tau}{x}{t}{\tau_1}{\tau_2}\\
%% %& & \pipe & 
%%%                           \tdencase{\tau}{y}{x}{\tau_1}{\tau_2}
%%%                            \\
%%
% & \pipe & \titerm{\iota}
%     \pipe \titype{\sigma} \pipe \trepof{\tau} \\
%   & \pipe &  (\Gamma; e)? \pipe \FF{syn}(\tau) \pipe \FF{ana}(\tau; \tau')\\
%%\textbf{external terms}         & e & ::= & \evar{x} \pipe 
%%%                           \efix{x}{\tau}{e} \pipe 
%%                            \elam{\evar{x}}{\tau}{e} \pipe 
%%                            \eop{tycon}{op}{
%%                              \tauidx
%%                            }{
%%                                  \splat{e}{1}{n}
%%                            } \\
%%                  &     &   &   \\
%%
%%
%%                        
%%%\text{deabstracted}& \iota & ::= & \mathcal{G}[\iota, \sigma]\\
%%                        
%%\\
%%\\
%%              
%%% &  & \pipe & \tvalof{\tau_1}{\tau_2} \pipe \iup{\tau} \\
%%%                         \trepof{\tau} \pipe \dup{\tau}\\
%%\text{translational IL} & \bar{\iota} & ::= & x \pipe \ifix{x}{\bar \sigma}{\bar \iota} 
%%  %\pipe \ilam{x}{\bar \sigma}{\bar \iota} \pipe \iapp{\bar \iota_1}{\bar \iota_2} 
%%  \pipe \cdots \pipe \itransof{\tau} \\           
%% & \bar{\sigma} & ::= & \darrow{\bar \sigma_1}{\bar \sigma_2} \pipe \cdots \pipe \dup{\tau} \pipe \trepof{\tau} \\
%%%\text{abstracted} & \sabs & ::= & \darrow{\sabs_1}{\sabs_2} \pipe \cdots \pipe \sabsrep{\tau}
%%                      \\
%\kappa & ::= & \karrow{\kappa_1}{\kappa_2} \pipe \klist{\kappa} \pipe \FF{L} \pipe \kunit \pipe \kpair{\kappa_1}{\kappa_2} \pipe \cdots \\
% & \pipe & \kTypeBlur \pipe \FF{TT} \pipe \kITerm \pipe \kIType \pipe \Q
%%                          \klabel \pipe
%%                          \klist{\kappa} \pipe
%%                        \kunit \pipe 
%%                        \kpair{\kappa_{1}}{\kappa_{2}} \\
%%                        &&\pipe&
%%                        \ksum{\kappa_1}{\kappa_2} \pipe
%%                        \kTypeBlur \pipe \kDen \pipe 
%%                        \kIType               
%%%\textbf{ops signature}     & \Theta  & ::= & \kOpEmpty \pipe \kOp{\Theta}{op}{\kappai}\\
%%%                                   &   &   & \\
%\\
%\\
%\iota  & ::= & \evar{x} \pipe 
%                       \ifix{\evar{x}}{\sigma}{\iota} \pipe
%                       \ilam{\evar{x}}{\sigma}{\iota} \pipe 
%                       \iapp{\iota_{1}}{\iota_{2}} 
%                       \\ & \pipe & \iunit \pipe \ipair{\iota_1}{\iota_2} \pipe \ifst{\iota} \pipe \isnd{\iota} \pipe \cdots \pipe \iup{\tau}
%%\\&\pipe&               \cdots \pipe  
%%                        n \pipe \iop{\iota_{1}}{\iota_{2}} \pipe \iIfEq{\iota_{1}}{\iota_{2}}{\dint}{\iota_{3}}{\iota_{4}} \\
%%& \pipe & 
%%                        \iunit \pipe
%%                        \ipair{\iota_{1}}{\iota_{2}} \pipe 
%%                        \ifst{\iota} \pipe
%%                        \isnd{\iota} 
%%\\&\pipe&                       
%%                         \iinl{\sigma_2}{\iota_1} \pipe \iinr{\sigma_1}{\iota_2} \\&\pipe & \icase{e}{x}{e_1}{x}{e_2} \\
%\\ \sigma  & ::= &    \darrow{\sigma_1}{\sigma_2} \pipe \dunit \pipe \dpair{\sigma_1}{\sigma_2} %\pipe \dint \pipe \dunit \pipe \dpair{\sigma_1}{\sigma_2} \pipe \dsum{\sigma_1}{\sigma_2}  
%\pipe \cdots \pipe \dup{\tau}
%\end{array}$$
%%\vspace{-10pt}
%\caption{\small Syntax of Core $\lam@\texttt{lang}$. Here, $x$ ranges over external and internal language variables, $\tvar{t}$ ranges over type-level variables, $\fvar{tycon}$ ranges over type constructor names and $\ell$ ranges over labels.
%\label{grammar}}
%\end{figure}
%\begin{figure}
%\small
%\begin{flalign}
%& \F{tycon}\fvar{record}~\F{of}\klist{\kpair{\FF{L}}{\kTypeBlur}}~\{\\
%& \quad \FF{iana}~\{\tlam{i}{\kpair{\klist{\kpair{\FF{L}}{\kTypeBlur}}}{\klist{\FF{L}}}}{\tlam{a}{\klist{\Q}}{\\& \quad\quad\tfold{\tvar{zip3}~\tfst{\tvar{i}}~\tsnd{\tvar{i}}~\tvar{a}}
%  {\titerm{\tunit}}{h}{t}{r}{
% \\&\quad\quad\quad\tifeq{\tfst{\tvar{first}~\tvar{h}}}{\tvar{second}~\tvar{h}}{\FF{L}}{
% \\&\quad\quad\quad\quad
% \F{let}~\tvar{x}:\kITerm~=~\FF{ana}(\tvar{third}~\tvar{h}; \tsnd{\tvar{first}~\tvar{h}})
%~\FF{in}
%\\ & \quad\quad\quad\quad \tfold{\tvar{t}}{\tvar{x}
%}{\_}{\_}{\_}{\titerm{(\iup{\tvar{x}},\iup{\tvar{r}})}}  }{\terr}
%%     \F{let}~\tvar{tt}:\FF{TT}=~\FF{syn}(\tsnd{\tsnd{\tvar{h}}}) \F{in}\\
%%\\     & \quad\quad\quad 
%}
%}}\}\\
%%& \quad \FF{isyn}~\{\tlam{i}{\klist{\FF{L}}}{\tlam{a}{\klist{\Q}}{
%%\\&\quad\quad \tfold{\tvar{zip2}~\tvar{i}~\tvar{a}}{\tden{\titerm{()}}{\ttype{record}{\tnil{{\kpair{\FF{L}}{\kTypeBlur}}}}}}{h}{t}{r}{
%%\\&\quad\quad\quad \F{let}~\tvar{htt}:\FF{TT}~=~\FF{syn}(\tsnd{\tvar{h}})
%%~\FF{in}
%%\\&\quad\quad\quad \F{let}~\tvar{hty}:\kTypeBlur~=~\ttypeof{\tvar{htt}}
%%\\&\quad\quad\quad \F{let}~\tvar{ty}:\kTypeBlur~=~\ttype{record}{(\tfst{\tvar{h}}, \tvar{hty})::\tnil{{\kpair{\FF{L}}{\kTypeBlur}}}}
%%~\FF{in}
%%\\&\quad\quad\quad \tfold{\tvar t}{\tden{\itransof{\tvar{htt}}}{\tvar{ty}}}{\_}{\_}{\_}{
%%\\&\quad\quad\quad\quad \tden{\titerm{(\iup{\itransof{\tvar{htt}}},\iup{\itransof{\tvar{r}}})}}{\tvar{ty}}}}
%%}}\}\\
%%& \quad \FF{esyn}~\{\tlam{i}{\FF{L}}{\tlam{a}{\klist{\Q}}{
%%  \tvar{decons1}~\tvar{a}~\tlam{ty}{\kTypeBlur}{\tlam{x}{\kITerm}{
%%\\&\quad\quad   \tfamcase{\tvar{ty}}{record}{sig}{
%%\\&\quad\quad\quad \tfold{\tvar{sig}}{\terr}{h}{t}{r}{
%%\\&\quad\quad\quad\quad \tifeq{\tfst{\tvar{h}}}{\tvar{i}}{\FF{L}}{\tfold{\tvar{t}}{\tden{\tvar{x}}{\tsnd{\tvar{h}}}}{\_}{t'}{\_}{
%%\\&\quad\quad\quad\quad\quad \tfold{\tvar{t'}}{\tden{\titerm{\ifst{\iup{\tvar{x}}}}}{\tsnd{\tvar{h}}}}{\_}{\_}{\tvar{r'}}{
%%\\&\quad\quad\quad\quad\quad\quad \tden{\titerm{\isnd{\iup{{\itransof{\tvar r}}}}}}{\tsnd{\tvar{h}}}}}\\&\quad\quad\quad\quad\hspace{-3px}}{\tvar{r}}
%%}\\&\quad\quad\hspace{-3px}}{\terr}
%%  }}
%%}}\\
%%& \quad \FF{rep}~\{\tlam{i}{\klist{\kpair{\FF{L}}{\kTypeBlur}}}{  \\& \quad \quad \tfold{\tvar{i}}{\titype{\dunit}}{s}{j}{r}{
%%    \tfold{\tvar{j}}{\trepof{\tsnd{\tvar{s}}}}{\_}{\_}{\_}{\\
%%    &\quad\quad\quad
%%    \titype{\dpair{\dup{\trepof{\tsnd{\tvar{s}}}}}{\dup{\tvar{r}}}}
%%    }
%%  }}\}\\
%%& \}; \\
%%& \F{let}\tvar{R} : \kTypeBlur = \ttype{record}{(\ell_1, \ttype{record}{\tnil{{\FF{L}\times\kTypeBlur}}}) :: \tnil{{\FF{L}\times\kTypeBlur}}}~\F{in}\\
%%&\F{let}id = \elam{x}{\tvar{R}}{\{\ell_1=~x\cdot\ell_1\} :: \fvar{record}}\\
%%&\F{let}triv = id[\{\ell_1 = \{\}\}]\cdot\ell_1
%%%&\F{let}one = \FF{intro}[\ell_\text{succ}]
%%%&\F{let}plus = \elam{x}{\tvar{nat}}{\elam{y}{\tvar{nat}}{\FF{elim}[()](x; y; \\
%%%& \quad \elam{r}{\tvar{nat}}{intro[\ell_\text{succ}](r) : \tvar{nat}}})}
%\end{flalign}
%\caption{The definition of the record type in  using the desugarings in Figure \ref{desugaring} and with the addition of simple let bindings for both type and term variables (not shown). Some type-level helper functions also omitted for concision.}
%
%\label{record-theory}
%\end{figure}
%We will now give a core typed lambda calculus, $\lam@\texttt{lang}$, that captures the semantics described in the previous sections. It is intended to make precise how active typechecking and translation works and how our mechanism relates to existing work on bidirectional typechecking, type-level computation and typed compilation while abstracting away from the details of Python's syntax and imposing a stronger type-level semantics that will allow us to state metatheoretic properties of interest. We will assume a fixed base for functions (providing the standard semantics of lambda functions) and target language, which we here call the \emph{internal language}.
%
%The syntax of $\lam@\texttt{lang}$ is shown in Figure \ref{grammar} and an example \emph{program} that defines a type constructor, $\fvar{record}$, with a semantics similar to that given in Listing \ref{record}\footnote{Technically, this defines labeled tuples, because the order of labels matters.}, using it to write an identity function and compute the value $triv$ (the empty record), is shown in Figure \ref{record-theory}.
%
%%In this section, we will develop an ``actively typed'' version of the simply-typed lambda calculus with simply-kinded type-level computation called $\lam@\texttt{lang}$. More specifically, the level of types, $\tau$, will itself form a simply-typed lambda calculus. \emph{Kinds} classify type-level terms in the same way that types conventionally classify expressions. Types become just one kind  of type-level value (which we will write $\kTypeBlur$, though it is also variously written $\star$, \lstinlinep{T} and \lstinlinep{Type} in various settings). Rather than there being a fixed set of type and operator constructors, we allow the programmer to declare new  constructors, and give their static and dynamic semantics by writing type-level functions. The kind system combined with techniques borrowed from the typed compilation literature and a form of type abstraction will allow us to prove strong type safety, decidability and conservativity theorems.
%%
%%The syntax of Core $\lam@\texttt{lang}$ is given in Fig. \ref{grammar}. An example of a program defining type and operator constructors that can be used to construct an active embedding of G\"odel's \textbf{T} into $\lam@\texttt{lang}$ is given in Fig. \ref{nat}. We will discuss its semantics and how precisely the embedding, seen being used starting on line 15 to ultimately compute the sum of two and two, works as we go on. Natural numbers can, of course, be isomorphically embedded in existing languages, with a similar usage and asymptotic performance profile (up to function call overhead as an abstract type, for example). We will provide more sophisticated examples where this is less feasible later on (and note that type abstraction is an orthogonal mechanism).
%%a type constructor declaration, $\fvar{nat}$, indexed trivially, together with three operator constructors, also all indexed trivially, that implement the standard introductory forms for natural numbers as well as the recursor operator (as in G\"odel's T \cite{pfpl}). Following the type constructor declaration, we apply $\fvar{nat}$ with the trivial index, $\tunit$, to form the type $\tvar{nat}$. Finally, we write an external term that uses the operators associated with $\fvar{nat}$ and the built-in constructor $\fvar{Parr}$, governing partial functions, to define an addition function and compute the addition of the natural numbers  two and two. We will introduce a more convenient concrete syntax in later portions of this thesis; for now we will restrict ourselves to the abstract syntax so that this example can directly aid in understanding the semantics.
%
%\subsection{Overview}\label{programs}
%
%A \emph{program}, $\rho$, consists of a series of constructor declarations followed by an external term, $e$.  Compiling a program consists of first \emph{kind checking} it (see below), then typechecking the external term and simultaneously {translating} it to a term, $\iota$, in the {typed internal language}. 
%%These correspond to the premises of the \emph{central compilation judgement} $\pkcompiles{\rho}{\iota}$:
%%\[
%%\inferrule[p-compiles]{
%%  \emptyset \vdash_{\fvalCtx_0} \rho\\
%%% \progOK{\emptyset}{\fvalCtx_0}{\rho}\\
%%  \pcompiles{\fvalCtx_0}{\rho}{\iota}
%%}{\pkcompiles{\rho}{\iota}}
%%\]
%%This anchors our exposition; we will describe how it is derived (i.e. how to write a compiler for $\lam@\texttt{lang}$) in the following sections. 
%The key judgements are the \emph{bidirectional active typing judgements}  (Fig. \ref{att}, which we describe starting in Sec. \ref{opcons}). They relate an external term, $e$, to a {type}, $\tau$, called its \emph{type assignment}, and an internal term, $\iota$, called its \emph{translation}, under \emph{typing context} $\Gamma$ and \emph{constructor context} $\Phi$. The first is synthesis, the second analysis.
%\[\Gamma \vdash_\fvalCtx e \Rightarrow \tau \leadsto \iota
%~~~~~~~
%\Gamma \vdash_\fvalCtx e \Leftarrow \tau \leadsto \iota\]
%%\[\ecompilesAX{e}{\tau}{\iota}\]
%
%The typing context, $\Gamma$, maps variables to types in essentially the conventional way (\cite{pfpl} contains the necessary background for this section). The constructor context, $\Phi$, tracks user-defined type  constructors. % Weakening and exchange of the constructor context closely related to the issue of conservativity that we will return to. Each constructor is identified by name, so there is no analog to contraction.
%
%The dynamic behavior of an external term is determined entirely by its translation to the internal language, which has a conventional operational semantics. This form of semantics can be seen as lifting into the language specification the first stage of a type-directed compiler like the TIL compiler for Standard ML \cite{tarditi+:til-OLD} and has some parallels to the Harper-Stone semantics for Standard ML, where external terms were also given meaning by elaboration from the EL to an IL \cite{Harper00atype-theoretic}. %
%
%In $\lam@\texttt{lang}$, the internal language (IL) provides partial functions (via the generic fixpoint operator of Plotkin's PCF) and simple product  types for the sake of our example. In practice, the internal language could be any typed  language with a specification for which type safety and decidability of typechecking have been satisfyingly determined. The internal type system serves as a ``floor'': guarantees that must hold for terms of any type (e.g. that out-of-bounds access to memory never occurs) must be maintained by the internal type system. User-defined constructors can enforce invariants stronger than those the internal type system maintains at particular types, however. Performance is also ultimately limited by the internal language and downstream compilation stages that we do not here consider (safe compiler extension has been discussed in previous work, e.g. \cite{conf/pldi/TatlockL10}).
%
%
%\subsection{Types and Type-Level Computation}\label{types}
%$\lam@\texttt{lang}$ supports, and makes extensive use of, simply-kinded type-level computation. Specifically, type-level terms, $\tau$, themselves form a typed lambda calculus. The classifiers of type-level terms are called \emph{kinds}, $\kappa$, to distinguish them from  \emph{types}. Types are  type-level values of kind $\kTypeBlur$. In @\texttt{lang}, the type-level language (together with the level of programs) is written in Python, which can be thought of as having a rather limited kind system (with one kind, \lstinlinep{dyn}). Here, we are able to more precisely discuss the kinds of values in the type-level language. 
%
%In most languages, types are formed by applying one of a collection of \emph{type constructors} to zero or more \emph{indices}. In $\lam@\texttt{lang}$, the situation is notionally similar. User-defined type constructors can be declared at the top of a program (or lifted to the top, in practice) using \textsf{tycon}. Each constructor in the program must have a unique name, written e.g. \fvar{record}.\footnote{We assume naming conflicts can be avoided by some extrinsic mechanism.} A type constructor must also declare an \emph{index kind}, $\kappaidx$. 
%
% To permit the embedding of interesting type systems, the type-level language includes several kinds other than $\kTypeBlur$. We lift several functional data structures to the type level: here, only unit ($\kunit$), binary products ($\kpair{\kappa_1}{\kappa_2}$) and lists ($\klist{\kappa}$), in addition to labels (introduced as $\ell$, possibly with a subscript, having kind \FF{L}).  Our record type constructor is indexed by a list of pairs of {labels}  and types (a signature, in essence; line 1). The type constructor $\fvar{arrow}$ is included in the initial constructor context, $\fvalCtx_0$ and has index kind $\kpair{\kTypeBlur}{\kTypeBlur}$.
% In practice, one could include a richer functional programming language and retain the spirit of the calculus, as long as it does not introduce general recursion at the type level. 
%  
%A type is introduced by applying a type constructor to an index of this kind, written $\ttype{Tycon}{\tauidx}$.  For example, the type of natural numbers is indexed trivially, so it is written $\ttype{Nat}{\tunit}$. We see a record type, abbreviated $\tvar{R}$, constructed on line 26. 
%
%The kind $\kTypeBlur$ also has an elimination form, $$\tfamcase{\tau}{Tycon}{x}{\tau_1}{\tau_2}$$ allowing the extraction of a type index by case analysis against a contextually-available type constructor. To a first approximation, one might think of type constructors as constructors of a built-in open \cite{conf/ppdp/LohH06}, $\kTypeBlur$, at the type-level. Like open datatypes, there is no notion of exhaustiveness so the default case is required for totality. We will see where this is used shortly.
%
%
%
%%We will write the kind of types as $\kTypeBlur$, though it is also written $\star$ or \lstinlinep{Type} in various similarly structured languages (see Sec. \ref{related-work}). 
%% Rather than there being a fixed set of type constructors, we allow the programmer to declare new type  constructors, and give the static and dynamic semantics of their associated operators, by writing type-level functions. In the semantics for this calculus, our kind system combined with techniques borrowed from the typed compilation literature and a form of type abstraction allow us to prove strong type safety, decidability and conservativity theorems.
%
%
%
%% and integers ($\dint$). We also include labels ($\klabel$), written in a slanted font, e.g. $\tlabel{myLabel}$, which are string-like values that only support comparison and play a distinguished role in the expanded syntax, as we will later discuss. Our first example, $\fvar{nat}$, is indexed trivially, i.e. by unit kind, $\kunit$, so there is only one natural number type, $\ttype{nat}{\tunit}$, but we will show examples of type constructors that are indexed in more interesting ways in later portions of this work. For example, $\fvar{LabeledTuple}$ has index kind $\klist{\kpair{\klabel}{\kTypeBlur}}$. 
% 
% Type constructors are not first-class; they do not themselves have arrow kind as in some kind systems (e.g.  \cite{watkins2008specifying}; Ch. 22 of \emph{PFPL} describes a related system \cite{pfpl}). The type-level language does, however, include total functions of arrow kind, written $\karrow{\kappa_1}{\kappa_2}$. Type constructor application can be wrapped in a type-level function to emulate a first-class or uncurried version of a type constructor for convenience (indeed, such a wrapper could be generated automatically, though we do not do so). 
%
%Two type-level terms of kind $\kTypeBlur$ are equivalent if they apply the same constructor, identified by name, to equivalent indices. Going further, we ensure that deciding type equivalence requires only checking for syntactic equality after normalization by imposing the restriction that equivalence at a type constructor's index kind must be decidable in this way. Our treatment of equivalence in the type-level language is thus quite similar to the treatment of term-level equality using ``equality types'' in a language like Standard ML.
%% A kind $\kappa$ is an  \emph{equality kind} if $\kEq{\kappa}$ can be derived (see appendix). 
%Conditional branching on the basis of equality at an equality kind can be performed in the type-level language. Equivalence at arrow kind is not decidable by our criteria, so type-level functions cannot appear within type indices. This also prevents general recursion from arising at the type level. Without this restriction, a type-level function taking a type as an argument could ``smuggle in'' a self reference as a type index, extracting it via case analysis (continuing our analogy to open datatypes, this is closely related to the positivity condition for inductive datatypes in total functional languages like Coq).% as maintaining the metatheoretic guarantee that typing respects type equivalence would impose a substantial burden in such a setting.% (a na\"ive approach to this would impose non-trivial extrinsic proof obligations onto extension developers that, unlike in others in this thesis, could threaten type safety).
%
%Every type constructor also defines a \emph{representation}, a type-level function that associates with every type an internal \emph{representation type} (analagous to \lstinlinep{trans_type} above). We will return to this after introducing the external forms and operator definitions.
%
%\subsection{Core External Forms and Desugaring}\label{opcons}
%The syntax for external terms (Figure \ref{grammar}) contains variables, $\lambda$ terms, three generalized introductory forms and a single generalized elimination form. The introductory forms are either unascribed, ascribed with a type or ascribed with a type constructor, as in our discussion of @\texttt{lang}. These generalized forms take a single a type-level value as an {index} and $n \geq 0$ arguments, which are other external terms. To better motivate this choice, we can give a purely syntactic desugaring of a Python-like syntax with labels to these forms, shown in Figure \ref{desugaring}. It is instructive to rewrite lines 27-28 of Figure \ref{record-theory} using these desugarings:% In @\texttt{lang}, desugarings can be user-defined 
%\noindent
%\vspace{-5px}
%$$
%\begin{array}{l}
%\FF{let}~id=\lambda x:\tvar{R}.\FF{intro}[\ell_1 :: []](x\cdot\FF{elim}[\ell_1]()) :: \fvar{record}\\
%\FF{let}~triv=id\cdot\FF{elim}[()](\FF{intro}[\ell_1 :: []](\FF{intro}[[]]()))\cdot\FF{elim}[\ell_1]()
%\end{array}
%$$
%\vspace{-5px}
%\begin{figure}[t]
%\small
%\[
%\begin{array}{rcl}
%\{ \ell_1{=}~e_1, \ldots, \ell_n{=}~e_n \} & := & \FF{intro}[\ell_1 :: \ldots :: \ell_n :: []](e_1; \ldots; e_n)\\
%(e_1, ..., e_n) & := & \FF{intro}[()](e_1; \ldots; e_n)\\
%\#n & := & \FF{intro}[n]()\\
%"s" & := & \FF{intro}[s]()\\
%e.\ell & := & e\cdot\FF{elim}[\ell]()\\
%e[e_1; \ldots; e_n] & := & e\cdot\FF{elim}[()](e_1; \ldots; e_n)\\
%e.\ell(e_1; ...; e_n) & := & e\cdot\FF{elim}[\ell](e_1; \ldots; e_n)
%%(e_1, ..., e_n) & := & intro[()](e_1; ...; e_n)
%%n &  := & intro[n]()
%%s & := & intro[s]()
%%
%%e.l & := & elim[l](e)
%%e[e_1; ...; e_n] & := & elim[()](e_1, ..., e_n)
%%e.l(e_1; ...; e_n) & := & elim[l]
%\end{array}
%\]
%\caption{Desugaring from conventional syntax to core forms. We assume that the type-level language has numbers, $n$, and strings, $s$ (not shown for concision).}
%\label{desugaring}
%\end{figure}
%
%%User-defined operator constructors are declared using \textsf{opcon}.  For reasons that we will discuss, our calculus associates every operator  constructor with a type constructor. The \emph{fully-qualified name} of every operator constructor, e.g. $\fvar{nat}.\opvar{z}$, must be unique. Operator constructors, like type constructors, declare an index kind, $\kappaidx$. In our first example, all the operator constructors are indexed trivially (by index kind $\kunit$), but other examples use more interesting indices. For example, in SML, the projection operator \lstinlinep{#3} can be applied to an $n$-tuple, $e$, iff $n \geq 3$. Note that it thus cannot be a function with a standard arrow type. Notionally, \lstinlinep{#} is an operator constructor and \lstinlinep{3} is its index. In an active embedding of $n$-tuples into $\lam@\texttt{lang}$, this would be written $\eop{Tuple}{prj}{3}{e}$ (we will nearly recover ML's syntax later). $\fvar{LabeledTuple}.\opvar{prj}$ is the operator constructor used to access a field of a labeled tuple, so it has index kind $\klabel$. An operator itself is, notionally, selected by indexing an operator constructor, e.g. $\fvar{nat}.\opvar{s}\langle \tunit \rangle$, but technically neither operator constructors nor operators are first-class at any level (additional machinery would be needed, e.g. an \textsf{Op} kind, but this is not fundamental to our calculus). 
%%Instead, in the external language, an operator constructor is applied by simultaneously providing an index and  $n \geq 0$ \emph{arguments}, written $\eop{Tycon}{op}{\tauidx}{\splat{e}{1}{n}}$\footnote{It may be helpful to distinguish between type/operator constructors and \emph{term formers}. There are term formers at all levels in the calculus. For example, operator constructor application and $\lambda$ are  external term formers, and type constructor application is a type-level term former. We might write these following Harper's conventions for abstract syntax to highlight this distinction \cite{pfpl}: $\mathtt{lam}[\tau](x.e)$, $\mathtt{ocapp}[\fvar{Tycon}, \opvar{op}, \tauidx](\splat{e}{1}{n})$ and $\texttt{tcapp}[\fvar{Tycon}](\tau)$.}. For example, on line 18 of Fig. \ref{nat}, we see the operator constructors $\fvar{nat}.\opvar{z}$ and $\fvar{nat}.\opvar{s}$ being applied to compute $two$. %\footnote{Although our focus here is entirely on semantics, a brief note on syntax: in the expanded syntax, the trivial indices and empty argument lists can be omitted, so we could write \texttt{Nat.s(Nat.s(Nat.z))}. With the ability to ``open'' a type's operators into the context, we could shorten this still to \texttt{s(s(z))}. Alternatively, with the ability to define a TSL in a manner similar to that in Sec. \ref{aparsing}, we might instead just write \texttt{2}.}
%
%\subsection{Operator Definitions and Representational Consistency}
%
%%
%\newcommand{\atjsynX}[3]{\Gamma \vdash_\fvalCtx #1 \Rightarrow #2 \leadsto #3}
%\newcommand{\atjanaX}[3]{\Gamma \vdash_\fvalCtx #1 \Leftarrow #2 \leadsto #3}
%\begin{figure}[t]
%\small
%$\fbox{\inferrule{}{\ecompilesAX{e}{\tau}{\iota}}}$
%~~~~$\Gamma ::= \emptyset \pipe \Gamma, x \Rightarrow \tau$
%\begin{mathpar}
%\inferrule[att-var]{
% x \Rightarrow \tau \in \Gamma
%}{
% \atjsynX{x}{\tau}{x}
%}
%
%\inferrule[att-syn-to-ana]{
% \atjsynX{e}{\tau}{\iota}
%}{
% \atjanaX{e}{\tau}{\iota}
%}
%
%\inferrule[att-lam]{
% \tau_1 \Downarrow_{\Gamma;\fvalCtx} \tau_1'\\
%   \trepof{\tau_1'} \Downarrow_{\Gamma;\fvalCtx} \titype{\sigma}\\\\
% \Gamma, x \Rightarrow \tau_1' \vdash_\fvalCtx e \Rightarrow \tau_2 \leadsto \iota
%%  \iota \hookrightarrow_\fvalCtx \iota'\\
%%    \sigma \hookrightarrow_\fvalCtx \sigma'
%%  \ddbar{\fvar{Arrow}}{\fvalCtx}{\trepof{\tau_1'}}{\sbar_1}\\
% %\delfromtau{$\Xi_0$}{\fvalCtx}{\tau_1'}{\sabs}\\\\
%}{
% \atjsynX{\elam{x}{\tau_1}{e}}{\ttype{arrow}{(\tau_1', \tau_2)}}{\ilam{x}{\sigma'}{\iota}}
%}
%
%\inferrule[att-i-unasc]{
% \vdash_\fvalCtx \FF{iana}(\fvar{tycon})=\taudef\\
% \taudef~(\tauidx, \tauidx')~((\Gamma; e_1)? :: \ldots :: (\Gamma; e_n)? :: []) \Downarrow_{\Gamma;\fvalCtx} \titerm{\iota}\\
%%  \trepof{\ttype{tycon}{\tauidx'}} \Downarrow_{\Gamma;\fvalCtx} \titype{\sigma}\\
%   \trepof{\ttype{tycon}{\tauidx'}} \Downarrow_{\Gamma;\fvalCtx} \titype{\sigma}\\
%   \Gamma \vdash_\fvalCtx \iota : \sigma
%}{
% \atjanaX{\FF{intro}[\tauidx](e_1; \ldots; e_n)}{\ttype{tycon}{\tauidx'}}{\iota}
%}
%
%\inferrule[att-i-asc-ty]{
% \atjanaX{I}{\tau}{\iota}
%}{
% \atjsynX{I : \tau}{\tau}{\iota}
%}
%
%\inferrule[att-i-asc-tycon]{
% \vdash_\fvalCtx \FF{isyn}(\fvar{tycon})=\taudef\\\\
% \taudef~\tauidx~((\Gamma; e_1)? {::}{\ldots}{::}(\Gamma; e_n)? {::} []) \Downarrow_{\Gamma;\fvalCtx} \tden{\titerm{\iota}}{\ttype{tycon}{\tauidx'}}\\
%     \trepof{\tau_1'} \Downarrow_{\Gamma;\fvalCtx} \titype{\sigma}\\
% \Gamma \vdash_\fvalCtx \iota : \sigma
%}{
% \atjsynX{\FF{intro}[\tauidx](e_1; \ldots; e_n)] :: \fvar{tycon}}{\tau}{\iota'}
%}
%
%\inferrule[att-elim]{
% \atjsynX{e}{\ttype{tycon}{-}}{-}\\
% \vdash_\fvalCtx \FF{esyn}(\fvar{tycon})=\taudef\\
% \taudef~((\Gamma; e)? :: (\Gamma; e_1)? :: \ldots :: (\Gamma; e_n)? :: []) \Downarrow_{\Gamma;\fvalCtx} \tden{\tau}{\iota}\\
%     \trepof{\tau} \Downarrow_{\Gamma;\fvalCtx} \titype{\sigma}\\
% \Gamma \vdash_\fvalCtx \iota : \sigma
%}{
% \atjsynX{e\cdot\FF{elim}[\tauidx](e_1; \ldots; e_n)}{\tau}{\iota}
%}
%\end{mathpar}
%\caption{\small The active typing judgement. The normalization judgement for type-level terms ($\Downarrow$) and the representational consistency check will be in an appendix.}
%\label{atj}
%\end{figure}
%\begin{figure}
%\small
%\begin{mathpar}
%\inferrule[repof]{
% \tau \Downarrow_{\Gamma;\fvalCtx} \ttype{tycon}{\tauidx}\\
% \vdash_\fvalCtx \FF{rep}(\fvar{tycon}) = \taurep\\
% \taurep \tauidx \Downarrow_{\Gamma;\fvalCtx} \titype{\sigma}
%}{
% \FF{repof}(\tau) \Downarrow_{\Gamma;\fvalCtx} \titype{\sigma}
%}
%
%\inferrule[syn]{
%    \tau \Downarrow_{\Gamma;\fvalCtx} (\Gamma; e)?\\
% \atjsynX{e}{\tau}{\iota}
%}{
% \FF{syn}(\tau) \Downarrow_{\Gamma;\fvalCtx} \tden{\titerm{\iota}}{\tau}
%}
%
%\inferrule[ana]{
% \tau_1 \Downarrow_{\Gamma;\fvalCtx} (\Gamma; e)?\\
% \tau_2 \Downarrow_{\Gamma;\fvalCtx} \ttype{tycon}{\tauidx}\\\\
% \atjanaX{e}{\ttype{tycon}{\tauidx}}{\iota}\\
%}{
% \FF{ana}(\tau_1; \tau_2) \Downarrow_{\Gamma;\fvalCtx} \titerm{\iota}
%}
%\end{mathpar}
%\caption{\small Evaluation semantics for the type-level language. Missing rules (including error propagation rules, which immediately cause failure of typechecking) are unsurprising and will be given in an appendix.}
%\label{tleval}
%\end{figure}
%The expressive and metatheoretic power of the calculus arises from how the rules for the active typing judgement handle these  generalized forms. Rather than fixing the specification of a finite collection of operator constructors and tasking the \emph{compiler} with deciding a typing derivation on its basis, the specification instead delegates to a type-level function associated with a type constructor. In the core calculus, this is one of three functions, called \FF{iana}, \FF{isyn} and \FF{esyn} in the grammar.\footnote{This means that any particular type constructor only supports a single introduction and elimination desugaring. This is a minor inconvenience in some cases that is resolved in @\texttt{lang} by the use of methods.} We see how this is done with the rules for the active typing judgements, given in Figure \ref{atj}.
%
%The first two rules are standard in bidirectional type systems: variables synthesize types and if a term synthesizes a type, it can be analyzed against a type (cf. Listing \ref{context}). Variables translate to variables, and if we are simply converting from synthesis to analysis, translation is not affected.
%
%The rule $\rulename{att-lam}$ must take into account the fact that, because we support type-level computation, the type annotation on the argument may not be in normal form. Thus, we evaluate it to normal form. Note that the kinding rules (not shown) will guarantee that, because $\tau_1$ is of kind $\kTypeBlur$, its normal form, $\tau'_1$, is of the form $\ttype{tycon}{\tauidx}$ for some normal $\tauidx$. To generate an appropriate internal type annotation in the translation, we need to compute the representation type of $\tau_1'$. This involves calling the representation function associated with its type constructor (rule \fvar{repof} in Figure \ref{tleval}). The form $\titerm{\sigma}$ is a \emph{quoted internal type}. Note in the syntax that there is an unquote form, $\dup{\tau}$, which allows us to compose internal types compositionally without needing to expose an elimination form. Indeed, it is an interesting facet of our calculus that we never need to examine syntax trees directly to implement extensions. The evaluation semantics remove quotations, so the normal form of a quoted internal type contains an internal type with no quotations.
%
%Lambda functions are the only introductory form requiring special support in the calculus (because they need to  manipulate the context; see Discussion). The next three rules show how any other abstractions that we define (e.g. records, decimals, etc.) make use of a generalized introductory form. 
%
%The rule  $\rulename{att-i-unasc}$ shows that unascribed introductory forms can only be analyzed against a type. Given such a type, the rule extracts a definition, named $\FF{iana}$, from the type constructor (cf. the \lstinlinep{ana_Dict} and \lstinlinep{trans_Dict} methods earlier). It calls this function with a pair containing the operator and type index and a list of \emph{reified arguments}, which have kind $\Q$ and introductory form $(\Gamma; e)?$. These are only constructed by the compiler (there would be no corresponding form in the concrete syntax). Their purpose is to allow the definition to programatically invoke synthesis and analysis as needed using the $\FF{ana}$ and $\FF{syn}$ operators. We can see the $\fvar{record}$ type constructor doing so on line 5 of Figure \ref{record-theory} to ensure that the field value provided as an argument has the same type as the corresponding label (accomplished by simultaneously folding over all three pieces of input data). The rule for performing analysis, $\fvar{ana}$, is in Figure \ref{tleval}. If it succeeds, analysis returns a \emph{translation}, which is a quoted internal term with kind $\kITerm$ and introductory form $\titerm{\iota}$. Like quoted internal types, there is an unquote form that is eliminated during evaluation. The operator definition uses this to construct a translation for the record. As before, empty records translate to units and records with a single field are unadorned. In this example, records with two or more fields translate to nested tuples. 
%
%Because we have a representation type associated with the type, we can check that the translation is \emph{representationally consistent} (the final premise). As we will discuss, representational consistency combined with type safety of the internal language implies type safety overall (it arises as a strengthening of the inductive hypothesis needed to prove type safety, and is closely related to work on \emph{type-preserving compilation} in the TIL compiler for Standard ML, \cite{tarditi+:til-OLD}).
%
%The next rule, $\rulename{att-i-asc-ty}$ states that introductory forms ascribed with a type are analyzed against that type.
%
%Introductory forms ascribed with a type constructor can, according to the final introductory rule, $\rulename{att-i-asc-tycon}$, synthesize a type via the definition $\FF{isyn}$. It is passed the operator index (there is no type index, since we only have a type constructor) and a list of arguments, as before. In this case, it must return not just a translation, but also a type. We use the form $\tden{\tau_2}{\tau_1}$ for such a pairing, which has kind $\FF{TT}$. The type and translation can be extracted from it using the appropriately named elimination forms (indeed, as given, it is merely a pair, but we give it special syntax for clarity -- it looks like the conclusion of the typing judgement -- and for other reasons that will become clear in future work). To synthesize a type from a list of labels and arguments, we must be able to synthesize types from the arguments. The $\FF{syn}(\tau)$ operator permits this, per rule $\fvar{syn}$, seen being used in Listing \ref{record-theory}. We check representational consistency of the result, as before.
%
%Finally, we show the rule for elimination forms. It operates by first synthesizing a type for the ``primary'' subterm, then extracting the $\FF{esyn}$ definition from its type constructor. As before, it is called with the arguments and representational consistency is checked. Note that the primary subterm is itself the first argument (though we have already synthesized a type for it, it is more uniform and clear to allow the operator  to do so again, which is done in our example by the $\tvar{decons1}$ helper function).
%
%\subsubsection{Representational Consistency}
%The representational consistency lemmas says that active typechecking and translation of an external term, if it succeeds (producing a type, $\hat\sigma$ and a translation $\iota$), always produces a well-typed translation. More specifically, the translation's type is the representation type of $\hat\sigma$.
%
%\begin{theorem}[Representational Consistency]
%Given well-kinded contexts and an external term that is both well-kinded and well-typed:
%\begin{enumerate}
%\item $\vdash \Phi$ 
%\item $\vdash_\Phi \Gamma$
%\item $\vdash_\Phi e$
%\item Either:
%\begin{enumerate}
%\item[a.] $\Gamma \vdash_\Phi e \Leftarrow \hat{\sigma} \leadsto \iota$
%\item[b.] $\Gamma \vdash_\Phi e \Rightarrow \hat{\sigma} \leadsto \iota$
%\end{enumerate}
%\end{enumerate}
%there exists an internal typing context, $\Omega$, and an internal type, $\tau$, such that  $\vdash_\Phi \Gamma \leadsto \Omega$ and $\FF{rep}(\hat\sigma) \Downarrow_\Phi \titype{\tau}$ and $\Omega \vdash \iota : \tau$.
%\end{theorem}
%\begin{proof}
%The proof proceeds by induction on the typing derivation (4a or 4b). Because the representational consistency is explicitly checked in the rules for intro and elim forms, these cases follow directly. The remaining cases are standard and follow inductively, given standard lemmas about valid contexts and a lemma about type safety of the static language. This is a fixed, standard functional language with only a few simple additions (TODO detail this).
%\end{proof}
%
%\subsubsection{Type Safety}
%Representational safety implies that well-typed external terms produce well-typed internal terms. If the internal language is type safe, then we know that well-typed external terms do not ``go wrong''. The internal language is simply PCF with sums and products, and a base type of integers, so this is a completely standard result \cite{pfpl}.

%\subsubsection{Decidability}
%The metatheoretic properties of interest are: type safety and termination of the type-level language (guaranteed by our kind system, though the details will need to be provided in an appendix; termination is non-trivial), type safety and decidability of the internal language (it i.s a standard variant of PCF, so this is trivial) and representational consistency. Type safety for the language as a whole comes as a corollary of these lemmas. We plan to provide detailed proofs of these, but for now, they should be treated as conjectures and our formulation above as expository. 
%We believe that despite this, the formalization is surprisingly elegant and concise, given its expressive power. It is a useful exercise to implement natural numbers, ala G\"odel's \textbf{T}, using this calculus, assuming that the type-level and internal languages include integers and one can lift them from the former into the latter.
%%
%
%For the synthetic functions, it must decide the type assignment and the translation on the basis of the index and arguments, or decide that this is not possible. An operator constructor with index kind $\kappaidx$ must have a definition of kind \[\kappaidx \rightarrow \klist{\kDen} \rightarrow (\kDen + \kunit)\] 
%As this suggests, the active typing judgement  will provide the operator index and a list of {recursively determined \emph{derivates}}, which have kind $\kDen$, for the arguments and ask the definition to return a value of ``option kind'', $\ksum{\kDen}{\kunit}$, where the trivial case indicates that a derivate cannot be constructed due to an invalid index, an incorrect number of arguments or an argument of an invalid type.\footnote{In practice, we would require operator constructor providers to report information about the precise location of the error (e.g. which argument was of incorrect type) and provide an appropriate error message and other metadata, but we omit this in the semantics.}
%
%
%A \emph{program}, $\rho$, consists of a series of type constructor declarations followed by an external term, $e$. The syntax for external terms contains six forms: variables, $\lambda$ terms, a generalized unascribed introductory form, a generalized introductory form ascribed with a type, a generalized introductory form ascribed with a type constructor and a generalized elimination form. We will discuss how a more natural syntax  like Python's can be added by simple desugaring.\todo{syntax}
%
%
%
%Compiling a program consists of first \emph{kind checking} it (typechecking the type-level terms, which has no analog in @\texttt{lang} because the type-level language is Python; discussed further below), then typechecking the external term and {translating} it to a term, $\iota$, in the {typed internal language}. %These correspond to the premises of the \emph{central compilation judgement} $\pkcompiles{\rho}{\iota}$:
%%\[
%%\inferrule[p-compiles]{%
%%  \emptyset \vdash_{\fvalCtx_0} \rho\\
%%  \progOK{\emptyset}{\fvalCtx_0}{\rho}\\
%%  \pcompiles{\fvalCtx_0}{\rho}{\iota}
%%}{\pkcompiles{\rho}{\iota}}
%%\]
%%his anchors our exposition; we will describe how it is derived (i.e. how to write a compiler for $\lam@\texttt{lang}$) in the following sections. 
%
%The key judgement in the calculus is the \emph{active typing judgement}  (Fig. \ref{att}, which we describe starting in Sec. \ref{opcons}). It relates an external term, $e$, to a {type}, $\tau$, called its \emph{type assignment}, and an internal term, $\iota$, called its \emph{translation}, under \emph{typing context} $\Gamma$ and \emph{constructor context} $\Phi$: 
%\[\ecompilesX{e}{\tau}{\iota}\]
%
%The typing context $\Gamma$ maps variables to types in essentially the conventional way (\cite{pfpl} contains the necessary background for this paper). The constructor context tracks user-defined type and operator constructors. % Weakening and exchange of the constructor context closely related to the issue of conservativity that we will return to. Each constructor is identified by name, so there is no analog to contraction.
%There is no separate operational semantics for the external language. Instead, the dynamic behavior of an external term is determined by its translation to the internal language, which has a more conventional operational semantics. This form of semantics can be seen as lifting into the language specification the first stage of a type-directed compiler like the TIL compiler for Standard ML \cite{tarditi+:til-OLD} and has some parallels to the Harper-Stone semantics for Standard ML, where external terms were also given meaning by elaboration from the EL to an IL \cite{Harper00atype-theoretic}. %
%
%In $\lam@\texttt{lang}$, the internal language (IL) provides partial functions (via the generic fixpoint operator of Plotkin's PCF), simple product and sum types and a base type of integers (to make our example interesting and as a nod toward speed on contemporary machines). In practice, the internal language could be any typed  language with a specification for which type safety and decidability of typechecking have been satisfyingly determined. The internal type system serves as a ``floor'': guarantees that must hold for \emph{all} well-typed terms, independent of the constructor context (e.g. that out-of-bounds access to memory never occurs), must be maintained by the internal type system. User-defined constructors can (only) enforce invariants stronger than those the internal type system maintains. Performance is also ultimately limited by the internal language and downstream compilation stages that we do not here consider (safe compiler extension has been discussed in previous work, e.g. \cite{conf/pldi/TatlockL10}).
%

%\begin{figure}
%\small
%\begin{flalign}
%\small
% & \family{Tuple}{\klist{\kTypeBlur}}{\\
% & \quad \tops{new}{\kunit}{\_}{a}{\tlam{\_}{\kunit}{\tlam{a}{\klist{\kDen}}{\\
% & \quad\quad
%   \tfold{\tvar{a}}{\tden{{()}}{\ttype{Tuple}{\tnil{\kTypeBlur}}}}{d}{b}{r}{\\
%     & \quad\quad\quad \tfamcase{\ttypeof{\tvar{r}}}{Ntuple}{i}{\\
%     & \quad\quad\quad\quad \tfold{\tvar{b}}{\tden{\itransof{\tvar{d}}}{\ttype{Ntuple}{\tcons{\ttypeof{\tvar{d}}}{\tvar{i}}}}\\& \quad\quad\quad\quad}{\_}{\_}{\_}{\tden{\\& \quad\quad\quad\quad\quad {(\itransof{\tvar{d}},\itransof{\tvar{r}})}}{\ttype{Ntuple}{\tcons{{\ttypeof{\tvar{d}}}}{\tvar{i}}}}}
%     }{\terr}
% }
% }}};\\
% & \quad \tops{prj}{\kint}{i}{a}{\lambda \tvar{i}:\kint.\lambda \tvar{a}:\klist{\kDen}. 
%   \tapp{\tvar{decons1}}{\tapp{\tvar{a}}{\tlam{d}{\kDen}{\\
% & \quad\quad\tfamcase{\ttypeof{\tvar{d}}}{Ntuple}{nl}{\\
% & \quad\quad \tfold{\tvar{nl}}{\terr}{t1}{j}{\_}{\\
% & \quad\quad \tfold{\tvar{j}}{\tifeq{\tvar{i}}{1}{\dint}{\tden{\itransof{d}}{\tvar{t1}}}{\terr}\\&\quad\quad\quad}{\_}{\_}{\_}{\\
% & \quad\quad\quad (\tlam{p}{\kpair{\kDen}{\dint}}{\tifeq{\tsnd{\tvar{p}}}{\tvar{i}}{\dint}{\tfst{\tvar{p}}}{\terr}})\\
% & \quad\quad\quad (\tapp{\tvar{foldl}}{\tapp{\tvar{nl}}{\tapp{(\tden{\tvar{x}}{\tvar{nt}},0)}{
%   \tlam{r}{\kpair{\kDen}{\dint}}{\tlam{t}{\kTypeBlur}{\tlam{ts}{\klist{\kTypeBlur}}{\\
% & \quad\quad\quad\quad \tifeq{\tvar{i}}{\tsnd{\tvar{r}}}{\dint}{\tvar{r}}{\tdencase{\tfst{\tvar{r}}}{rx}{\_}{\\
% & \quad\quad\quad\quad\quad \tifeq{\tvar{i}}{\tsnd{\tvar{r}}+1}{\dint}{\\
% & \quad\quad\quad\quad\quad\quad \tifeq{\tvar{ts}}{\tnil{\kTypeBlur}}{\klist{\kTypeBlur}}{(\tden{\tvar{rx}}{\tvar{t}},\tvar{i})\\
% & \quad\quad\quad\quad\quad\quad}{(\tden{\titerm{\ifst{\iup{\tvar{rx}}}}}{\tvar{t}},\tvar{i})}\\  
% & \quad\quad\quad\quad\quad}{(\tden{\titerm{\isnd{\iup{\tvar{rx}}}}}{\tvar{t}},\tsnd{\tvar{r}}+1)}}{\terr}}}}}
% }}})
% }}{\terr}}}}}
% }\\ & 
%%& \quad \tops{pr}a}{b}{c}{d}
% }{d}{i}{(\tlam{i}{\klist{\kTypeBlur}}{
%   \\& \quad \quad \tfold{\tvar{i}}{\titype{\dunit}}{s}{j}{r}{
%     \tfold{\tvar{j}}{\titype{\trepof{\tvar{s}}}}{\_}{\_}{\_}{\\
%   &\quad\quad\quad
%     \titype{\dpair{\trepof{\tvar{s}}}{\dup{\tvar{r}}}}
%     }
%   }
% })}
%\end{flalign}
%\caption{ABC}
%\label{tuple}
%\end{figure}
%\begin{figure}[t]
%\small
%$\fbox{\inferrule{}{\progOKX{\progsort}}}$
%~~~$\tvarCtx ::= \emptyctx \pipe \tvarCtxX{t}{\kappa}$
%~~~$\fvalCtx ::= \emptyctx \pipe \fvalCtxX{\fvalDf}$
%%$\fCtx ::= \Sigma_0  \pipe \fvalCtxX$
%\begin{mathpar}
%\inferrule[k-tycon]{
% \fvar{tycon} \notin \text{dom}(\fvalCtx)\\
% \kEq{\kappaidx}\\
% \tKind{\tvarCtx}{\fvalCtx}{\taurep}{\karrow{\kappaidx}{\kIType}}\\
% \opType{\tvarCtx}{\fvalCtxX{\fvalDf}}{\theta}{\Theta}\\
% \progOK{\tvarCtx}{\fvalCtxX{\fvalDf}}{\rho}
%}{
% \progOKX{\pfam{\familyDf}{\rho}}
%}
%
%%\inferrule[def-kinding]{
%%  \tKindX{\tau}{\kappa}\\
%%  \progOK{\tvarCtxX{t}{\kappa}}{\fvalCtx}{\rho}
%%}{
%%  \progOKX{\pdef{t}{\kappa}{\tau}{\rho}}
%%}
%%
%\inferrule[k-e-prog]{
% \exprOK{\tvarCtx}{\fvalCtx}{e}
%}{
% \progOKX{e}
%}
%\end{mathpar}
%$\fbox{$\opType{\tvarCtx}{\fvalCtx}{\theta}{\Theta}$}$
%\begin{mathpar}
%\inferrule[k-opcon]{
% \tKind{\tvarCtx}{\fvalCtx}{\taudef}{\karrow{\kappaidx}{\karrow{\klist{\kDen}}{(\ksum{\kDen}{\kunit})}}}
%}{
% \opType{\tvarCtx}{\fvalCtx}{{\tops{op}{\kappaidx}{i}{a}{\taudef}}}{
% \kOpS{op}{\kappaidx}}
%}
%
%\inferrule[k-opcons]{
% \opType{\tvarCtx}{\fvalCtx}{\theta_1}{\Theta_1}\\
% \opType{\tvarCtx}{\fvalCtx}{\theta_2}{\Theta_2}\\\\
% \text{dom}(\theta_1) \cap \text{dom}(\theta_2) = \emptyset
%}{
% \opType{\tvarCtx}{\fvalCtx}{\theta_1; \theta_2}{\Theta_1, \Theta_2}
%}
%\end{mathpar}
%$\fbox{$\exprOKX{e}$}$
%\begin{mathpar}
%\inferrule[k-e-var]{ }{
% \exprOK{\tvarCtx}{\fvalCtx}{\evar{x}}
%}
%
%\inferrule[k-e-lam]{
% \tKindX{\tau}{\kTypeBlur}\\
% \exprOK{\tvarCtx}{\fvalCtx}{e}
%}{
% \exprOKX{\elam{\evar{x}}{\tau}{e}}
%}
%
%\inferrule[k-e-op]{
% \fval{tycon}{-}{-}{\theta} \in \fvalCtx\\
% \tops{op}{\kappaidx}{i}{a}{-} \in \theta\\
% \tKindX{\tauidx}{\kappaidx}\\
% \exprOKX{e_1}\\
% \cdots\\
% \exprOKX{e_n}
%}{
% \exprOKX{\eop{tycon}{op}{\tauidx}{\splat{e}{1}{n}}}
%}
%\end{mathpar}
%$\fbox{$\kEq{\kappa}$}$
%\begin{mathpar}
%\inferrule[t-eq]{ }{
% \kEq{\kTypeBlur}
%}
%
%\inferrule[i-eq]{ }{
% \kEq{\kint}
%}
%
%\inferrule[l-eq]{ }{
% \kEq{\klabel}
%}
%
%\inferrule[list-eq]{
% \kEq{\kappa}
%}{
% \kEq{\klist{\kappa}}
%}
%\\
%\inferrule[u-eq]{ }{
% \kEq{\kunit}
%}
%
%\inferrule[p-eq]{
% \kEq{\kappa_1}\\
% \kEq{\kappa_2}
%}{
% \kEq{\kpair{\kappa_1}{\kappa_2}}
%}
%
%\inferrule[s-eq]{
% \kEq{\kappa_1}\\
% \kEq{\kappa_2}
%}{
% \kEq{\ksum{\kappa_1}{\kappa_2}}
%}
%\end{mathpar}
%\caption{\small Kinding for type and operator constructors and external terms, and equality kinds (see text).  Type-level terms stored in the constructor context are not needed during kinding, indicated using a dash.}
%\label{kindprog}
%\vspace{-10pt}
%\end{figure}

%\section{Interactive Compilation and Invocation}\label{targets}
%\subsection{OpenCL as an Active Library}
%The code in this section uses \lstinlinep{clx}, an example library implementing the semantics of the OpenCL programming language and extending it with some additional useful types, which we will discuss shortly. @\texttt{lang} itself has no built-in support for OpenCL.
%
%To briefly review, OpenCL provides a data-parallel SPMD programming model where developers define functions, called {\em kernels}, for execution on \emph{compute devices} like GPUs or multi-core CPUs \cite{opencl11}. Each thread executes the same kernel but has access to a unique index, called its \emph{global ID}. Kernel code is written in a variant of C99 extended with some new primitive types and operators, which we will introduce  as needed in our examples below.
%
%\subsection{Generic Functions}\label{genfn}
%%\begin{codelisting}[t]
%%\lstinputlisting[commentstyle=\color{mauve}]{listing7.py}
%%\caption{[\texttt{listing\ref{metaprogramming}.py}] Metaprogramming with @\texttt{lang}, showing how to construct generic functions from abstract syntax trees.}
%%\label{metaprogramming}
%%\end{codelisting}
%
%%\begin{codelisting}
%%\lstinputlisting[linebackgroundcolor={\btLstHL{5-8}}]{listing3.py}
%%\caption{[\texttt{listing\ref{map}.py}] A generic data-parallel higher-order map function targeting OpenCL.}
%%\label{map}
%%\end{codelisting}
%%\begin{codelisting}[t]
%%\lstinputlisting[commentstyle=\color{mauve}]{listing7.py}
%%\caption{[\texttt{listing\ref{metaprogramming}.py}] Metaprogramming with @\texttt{lang}, showing how to construct generic functions from abstract syntax trees.}
%%\label{metaprogramming}
%%\end{codelisting}
%Lines 3-4 introduce \lstinlinep{map}, an @\texttt{lang} function of three arguments that is governed by the \emph{active base} referred to by \lstinlinep{clx.base} and targeting the \emph{active target} referred to by \lstinlinep{clx.opencl}. The active target determines which language the function will compile to (here, the OpenCL kernel language) and mediates code generation. 
%
%The body of this function, highlighted in grey for emphasis, does not have Python's semantics. Instead, it will be governed by the active base together with the active types used within it. No such types have been provided explicitly, however. Because our type system is extensible, the code inside could be meaningful for many different assignments of types to the arguments. We call functions awaiting types \emph{generic functions}. Once types have been assigned, they are called \emph{concrete functions}.
%
%Generic functions are represented at compile-time as instances of \lstinlinep{ace.GenericFn} and consist of an abstract syntax tree, an {active base} and an {active target}. The purpose of the \emph{decorator} on line 3 is to replace the Python function on lines 4-8 with an @\texttt{lang} generic function having the same syntax tree and the provided active base and active target. 
%Decorators in Python are simply syntactic sugar for applying the decorator function directly to the function  being decorated \cite{python}. In other words, line 3 could be replaced by inserting the following  statement on line 9:
%\vspace{-0.28cm}
%\begin{verbbox}
%map = ace.fn(clx.base, clx.opencl)(map)
%\end{verbbox}
%\begin{figure}[h!]
%\centering
%\theverbbox
%\end{figure}
%\vspace{-0.28cm}
%
%The abstract syntax tree for \lstinlinep{map} is extracted using the Python standard  library packages  \lstinlinep{inspect} (to retrieve its source code) and \lstinlinep{ast} (to parse it into a syntax tree). 
%
%%\subsection{Metaprogramming in @\texttt{lang}}
%%Generic functions can be generated directly from ASTs as well, providing @\texttt{lang} with support for  straightforward metaprogramming. Listing \ref{metaprogramming} shows how to generate two more generic functions, \lstinlinep{scale} and \lstinlinep{negate}. The latter is derived from the former by using a library for manipulating Python syntax trees, \lstinlinep{astx}. In particular, the \lstinlinep{specialize} function replaces uses of the second argument of \lstinlinep{scale} with the literal \lstinlinep{-1} (and changes the function's name), leaving a function of one argument.
% 
%\subsection{Concrete Functions and Explicit Compilation}
%To compile a generic function to a particular \emph{concrete function}, a type must be provided for each argument, and typechecking and translation must then succeed. Listing \ref{compscript} shows how to explicitly provide type assignments to \lstinlinep{map} using the subscript operator (implemented using Python's operator overloading mechanism). We attempt to do so three times in Listing \ref{compscript}. The first, on line \ref{compscript}.7, fails due to a type error, which we handle so that the script can proceed. The error occurred  because the ordering of the argument types was incorrect. We provide a valid ordering on line \ref{compscript}.9 to generate the concrete function \lstinlinep{map_neg_f32}. We then provide a different type assignment to generate the concrete function \lstinlinep{map_neg_ci32}.
%Concrete functions are instances of \lstinlinep{ace.TypedFn}, consisting of an abstract syntax tree annotated with types and translations along with a reference to the original generic function. %The typing and translation process is mediated by the logic in the active base, types and target that have been provided, as we will describe in more detail below. 
%
%To produce an output file from an @\texttt{lang} ``compilation script'' like \lstinlinep{listing}\texttt{\ref{compscript}}\lstinlinep{.py}, the command \lstinlinep{acec} can be invoked from the shell, as shown in Listing \ref{mapc}. 
%\subsection{Types}
%\begin{codelisting}
%\lstinputlisting{listing4.py}
%\caption{[\texttt{listing\ref{compscript}.py}] The generic \texttt{map} function compiled to map the \texttt{negate} function over two  types of input.}
%\label{compscript}
%\end{codelisting}
%\begin{codelisting}
%\begin{lstlisting}[style=Bash]
%$ `acec listing3.py`
%Hello, compile-time world!
%[ace] TypeError in listing1.py (line 6, col 28): 
%      'GenericFnType(negate)' does not support [].
%[acec] listing3.cl successfully generated.
%\end{lstlisting}
%\caption{Compiling \texttt{listing\ref{compscript}.py} using the \texttt{acec} compiler.}
%\label{mapc}
%\end{codelisting}
%\begin{codelisting}
%\lstinputlisting[style=OpenCL]{listing5.cl}
%\caption{[\texttt{listing\ref{compscript}.cl}] The OpenCL file generated by Listing \ref{mapc}.}
%\label{mapout}
%\end{codelisting}
%%Lines \ref{compscript}.3-\ref{compscript}.5 construct the types assigned to the arguments of \lstinlinep{map} on lines \ref{compscript}.7-\ref{compscript}.10. In @\texttt{lang}, types are themselves values that can be manipulated at compile-time. This stands in contrast to other contemporary languages, where user-defined types (e.g. datatypes, classes, structs) are written declaratively at compile-time but cannot be constructed, inspected or passed around programmatically. More specifically, types are instances of a Python class that implements the \lstinlinep{ace.ActiveType} interface (see Sec. \ref{atypes}). 
%%As Python values, types can be assigned to variables when convenient (removing the need for  facilities like \lstinlinep{typedef} in C or \lstinlinep{type} in Haskell). Types, like all compile-time objects derived from @\texttt{lang} base classes, do not have visible state and operate in a referentially transparent manner (by constructor memoization, which we do not detail here).% These types are all implemented in the \lstinlinep{clx} library imported on line 1, none are built into @\texttt{lang} itself.
%
%The type named \lstinlinep{T1} on line \ref{compscript}.3 corresponds to the OpenCL type \lstinlinep{global float*}: a pointer to a 32-bit floating point number stored in the compute device's global memory (one of four address spaces defined by OpenCL \cite{opencl11}). It is constructed by applying \lstinlinep{clx.Ptr}, which is an @\texttt{lang} type constructor corresponding to pointer types, to a value representing the  address space, \lstinlinep{clx.global_}, and the type being pointed to. That type, \lstinlinep{clx.float}, is in turn the @\texttt{lang} type corresponding to \lstinlinep{float} in OpenCL (which, unlike C99, is always 32 bits). 
%The \lstinlinep{clx} library contains a full implementation of the OpenCL type system (including behaviors, like promotions, inherited from C99).
%@\texttt{lang} is \emph{unopinionated} about issues like memory safety and the wisdom of such promotions. We will discuss how to implement, as libraries, abstractions that are higher-level than raw pointers in Sec. \ref{examples}, but @\texttt{lang} does not prevent users from choosing a low level of abstraction or ``interesting'' semantics if the need arises (e.g. for compatibility with existing libraries; see the discussion in Sec. \ref{discussion}). We also note that we are being more verbose than necessary for the sake of pedagogy. The \lstinlinep{clx} library includes more concise shorthand for OpenCL's types: \lstinlinep{T1} is equal to \lstinlinep{clx.gp(clx.f32)}. %Similarly, the decorators in Listings \ref{map} and \ref{metaprogramming} could have been written \lstinlinep{clx.cl_fn}.\todo{move this back there probably}
%
%The type \lstinlinep{T2} on line \ref{compscript}.4 is a pointer to a \emph{complex integer} in global memory. It does not correspond direrctly to a type in OpenCL, because OpenCL does not include primitive support for complex numbers. Instead, it uses an active type constructor \lstinlinep{clx.Cplx}, which includes the necessary logic for typechecking operations on complex numbers and translating them to OpenCL (Sec. \ref{atypes}). This constructor is parameterized by the numeric type that should be used for the real and imaginary parts, here \lstinlinep{clx.int}, which corresponds to 32-bit OpenCL integers. Arithmetic operations with other complex numbers, as well as with plain numeric types (treated as if their imaginary part was zero), are supported. When targeting OpenCL, @\texttt{lang} expressions assigned type \lstinlinep{clx.Cplx(clx.int)} are compiled to OpenCL expressions of type \lstinlinep{int2}, a  \emph{vector type} of two 32-bit integers (a type that itself is not inherited from C99). This can be observed in several places on lines \ref{mapout}.14-\ref{mapout}.21. This choice is merely an implementation detail that can be kept private to \lstinlinep{clx}, however. An @\texttt{lang} value of type \lstinlinep{clx.int2} (that is, an actual OpenCL vector) \emph{cannot} be used when a \lstinlinep{clx.Cplx(clx.int)} is expected (and attempting to do so will result in a static type error): \lstinlinep{clx.Cplx} truly extends the type system, it is not a type alias.
%
%The type \lstinlinep{TF} on line \ref{compscript}.5 is extracted from the generic function \lstinlinep{negate} constructed in Listing \ref{metaprogramming}. Generic functions, according to Sec. \ref{genfn}, have not yet had a type assigned to them, so it may seem perplexing that we are nevertheless assigning a type to \lstinlinep{negate}. Although a conventional arrow type cannot be assigned to \lstinlinep{negate}, we can give it a \emph{singleton type}: a type that simply means ``this expression is the \emph{particular} generic function \lstinlinep{negate}''. This type could also have been explicitly written as \lstinlinep{ace.GenericFnType(listing2.negate)}. During typechecking and translation of \lstinlinep{map_neg_f32} and \lstinlinep{map_neg_ci32}, the call to \lstinlinep{f} on line \ref{map}.6 uses the type of the provided argument to compile the generic function that inhabits the singleton type of \lstinlinep{f} (\lstinlinep{negate} in both of these cases) to a concrete function. This is why there are two versions of \lstinlinep{negate} in the output in Listing \ref{mapout}. In other words, types \emph{propagate} into generic functions -- we didn't need to compile \lstinlinep{negate} explicitly. This also explains the error printed on line \ref{mapc}.3-\ref{mapc}.4: when this type was inadvertently assigned to the first argument \lstinlinep{input}, the indexing operation on line \ref{map}.6 resulted in an error. A generic function can only be \emph{statically} indexed by a list of types to turn it into a concrete function, not \emph{dynamically} indexed with a value of type \lstinlinep{clx.size_t} (the return type of the OpenCL primitive function \lstinlinep{get_global_id}).
%
%In effect, this scheme enables higher-order functions even when targeting languages, like OpenCL, that have no support for higher-order functions (OpenCL, unlike C99, does not support function pointers). Interestingly, because they have a singleton type, they are higher-order but not first-class functions. That is, the type system would prevent you from creating a heterogeneous list of generic functions. Concrete functions, on the other hand, can be given both a singleton type and a true function type. For example, \lstinlinep{listing2.negate[[clx.int]]} could be given type \lstinlinep{ace.Arrow(clx.int, clx.int)}. The base determines how to convert the @\texttt{lang} arrow type to an arrow type in the target language (e.g. a function pointer for C99, or an integer that indexes into a jump table constructed from knowledge of available functions of the appropriate type in OpenCL).
%
%Type assignment to generic functions is similar in some ways to template specialization in C++. In effect, both a template header and type parameters at call sites are being generated automatically by @\texttt{lang}. This simplifies a sophisticated feature of C++ and enables its use with other targets like OpenCL. %Other uses for C++ templates (and the preprocessor) are subsumed by the metaprogramming features discussed above\todo{cite template metaprogramming paper}.
%
%%\subsection{Within-Function Type Resolution}
%%\begin{codelisting}
%%\lstinputlisting{listing6.py}
%%\caption{\texttt{[listing6.py]} A function demonstrating whole-function type inference when multiple values with differing types are assigned to a single identifier, \texttt{y}.}
%%\label{inference}
%%\end{codelisting}
%%On line 5 in the generic \lstinlinep{map} function in Listing \ref{map}, the variable \lstinlinep{gid} is initialized with the result of calling the OpenCL primitive \lstinlinep{get_global_id}.  The type for \lstinlinep{gid} is never given explicitly. This is a simple case of @\texttt{lang}'s {\em within-function type resolution} strategy (we hesitate to call it \emph{type inference} because it does not, strictly speaking, take a constraint-solving approach). In this case, the type of \lstinlinep{gid} will resolve to \lstinlinep{size_t} because that is the return type of \lstinlinep{get_global_id} (as defined in the OpenCL specification, which the \lstinlinep{ace.OpenCL} module follows). The result can be observed on Lines 11 and 24 in Listing \ref{mapout}. 
%%
%%Inference is not restricted within single assignments, as in the \lstinlinep{map} example, however. Multiple assignments to the same identifier with values of differing types, or multiple return statements, can be combined if the types in each case are compatible with one another (e.g. by a subtyping relation or an implicit coercion). In Listing \ref{inference}, the \lstinlinep{threshold_scale} function assigns different values to \lstinlinep{y} in each branch of the conditional. In the first branch, the value \lstinlinep{0} is an \lstinlinep{int} literal. However, in the second branch of the loop, the type depends on the types of both arguments, \lstinlinep{x} and \lstinlinep{scale}. We show two choices for these types on Lines 11 and 12. Type inference correctly combines these two types according to OpenCL's C99-derived rules governing numeric types (defined by the user in the \lstinlinep{OpenCL} module, as we will describe in Section \ref{att}). We can verify this programmatically on Lines 12 and 13. Note that this example would also work correctly if the assignments to \lstinlinep{y} were replaced with \lstinlinep{return} statements (in other words, the return value of a function is treated as an assignable for the purpose of type inference).
%
%\subsection{Implicit Compilation and Interactive Execution}\label{compenv}\label{backend}\label{implicit}
%%Professional end-user programmers (e.g. scientists and engineers \cite{professional-end-users}) today generally use dynamically-typed high-level languages like MATLAB, Python, R or Perl for tasks that are not performance-sensitive, such as small-scale data analysis and plotting \cite{nguyen2010survey}. For portions of their analyses where the performance overhead of dynamic type checking and automatic memory management is too high, they will typically call into code written in a statically-typed, low-level language, most commonly C or Fortran, that uses low-level parallel abstractions like pthreads and MPI \cite{4222616,basili2008understanding}. Unfortunately, these low-level languages and abstractions are notoriously difficult to use and automatic verification is intractable in general.\todo{integrate this}
%
%
%A common workflow for \emph{professional end-user programmers} (e.g. scientists and engineers) is to use a simple scripting language for orchestration, small-scale data analysis and visualization and call into a low-level language for performance-critical sections. Python is both designed for this style of use and widely adopted for such tasks \cite{sanner1999python,nguyen2010survey}. Developers can call into native functions using Python's foreign function interface (FFI), for example. A more recent trend is to generate and compile code without leaving Python, using a Python wrapper around a compiler. For example, \lstinlinep{weave} works with C and C++, and \lstinlinep{pycuda} and \lstinlinep{pyopencl} work with CUDA and OpenCL, respectively \cite{klockner2011pycuda}. 
%\begin{codelisting}
%\lstinputlisting{listing8.py}
%\caption{[\texttt{listing\ref{py}.py}] A full OpenCL program using the \texttt{clx} Python bindings, including data transfer to and from a device and direct invocation of a generic function, \texttt{map}.}
%\label{py}
%\end{codelisting}
%The OpenCL language was designed for this workflow, exposing a retargetable compiler and data management  routines as an API, called the \emph{host API} \cite{opencl11}. The \lstinlinep{pyopencl} library exposes this API to Python and simplifies interoperation with \lstinlinep{numpy}, a popular package for manipulating contiguously-allocated numeric arrays in Python \cite{klockner2011pycuda}.
%
%@\texttt{lang} supports a refinement to this workflow, as an alternative to the \lstinlinep{acec} compiler described above, for targets that have wrappers like this available, including \lstinlinep{clx.opencl}. Listing \ref{py} shows an example of this workflow where the user chooses a compute device (line \ref{py}.3),  constructs a \lstinlinep{numpy} array (line \ref{py}.5), transfers it to the device (line \ref{py}.6), allocates an empty equal-sized buffer for the result of the computation (line \ref{py}.7), launches the {generic} kernel \lstinlinep{map} from Listing \ref{map} with these device arrays as well as the function \lstinlinep{negate} from Listing \ref{metaprogramming} (line \ref{py}.9) choosing a number of threads equal to the number of elements in the input array (line \ref{py}.10), and transfers the result back into main memory to check that the device computed the expected result (line \ref{py}.12).
%
%For developers experienced with the usual OpenCL or CUDA workflow, the fact that this can be accomplished in a total of 6 statements may be surprising. This simplicity is made possible by @\texttt{lang}'s implicit tracking of types throughout the code. First, \lstinlinep{numpy} keeps track of the type, shape and order of its arrays. The type of \lstinlinep{input}, for example, is \lstinlinep{numpy.float64} by default, its shape is \lstinlinep{(1024,1024)} and its order is row-major by default. The \lstinlinep{pyopencl} library, which the \lstinlinep{clx} mechanism is built upon, uses this metadata to automatically call the underlying OpenCL host API function for transferring byte arrays to the device without requiring the user to calculate the size. The \lstinlinep{clx} wrapper further retains this metadata in \lstinlinep{d_input} and \lstinlinep{d_output}, the Python wrappers around the allocated device arrays. The @\texttt{lang} active type of these wrappers is an instance of \lstinlinep{clx.NPArray} parameterized by this metadata. This type knows how to typecheck and translate operations, like indexing with a multi-dimensional thread index, automatically.
%
%By extracting the types from the arguments, we can call the generic function \lstinlinep{map} without first requiring an explicit type assignment, like we needed when using \lstinlinep{acec} above. In other words, dynamic types and other metadata can propagate from Python data structures into an @\texttt{lang} generic function as static type information, in the same manner as it propagated \emph{between} generic functions in the previous section. In both cases, typechecking and translation of \lstinlinep{map} happens the first time a particular type assignment is encountered and cached for subsequent use. When called from Python, the generated OpenCL source code is compiled for the device we selected using the OpenCL retargetable compilation infrastructure, and cached for subsequent calls. 
%
%The same program written using the OpenCL C API directly is an order of magnitude longer and significantly more difficult to comprehend. OpenCL does not support higher-order functions nor is there any way to write \lstinlinep{map} in a type-generic or shape-generic manner. If we instead use the \lstinlinep{pyopencl} library and apply the techniques described in \cite{klockner2011pycuda}, the program is still twice as large and less readable than this code. Both the \lstinlinep{map} and \lstinlinep{negate} functions must be explicitly specialized with the appropriate types using string manipulation techniques, and custom shapes and orders can be awkward to handle. Higher order functions are still not available, and must also be simulated by string manipulation. That approach also does not permit the use any of the language extensions that @\texttt{lang} enables (beyond the useful type for \lstinlinep{numpy} arrays just described; see Sec. \ref{examples} for more interesting possibilities).
%
%%Not shown are several additional conveniences, such as delegated kernel sizing and \lstinlinep{In} and \lstinlinep{Out} constructs that can reduce the size and improve the clarity of this code further; due to a lack of space, the reader is referred to the language documentation.

\section{Related Work}\label{related}

%\parindent0pt
%
%\subsection{Active Libraries}

%We have implemented the full OpenCL programming language as  primary example showed how to support low-level GPU programming from a high-level language. Other languages have supported similar workflows, e.g. Rust \cite{rustgpu}. These have not been fundamentally extensible and suffer from the ``bootstrapping'' problem described in Sec. \ref{intro}.

Early work proposing the inclusion of compile-time logic in libraries led to the phrase {\it active libraries} \cite{activelibraries}. We borrow the prefix ``active''. Our focus on type constructors and type system extension differs substantially from previous work, which focused on term rewriting for the purpose of optimization over a fixed semantics. Several contemporary projects have the same goals, e.g. LMS \cite{Rompf:2012:LMS}, which allows staged translation of well-typed Scala programs to other targets and supports composing optimizations. Macro systems can also be used to define optimizations and operations with interesting dynamic semantics. In both cases, the language's type system itself remains fixed. Our mechanism permits true type system extensions. New types are not merely aliases, nor must their rules be admissible in some base type system (e.g. \lstinlinep{string_in} would be difficult to define in Scala, particularly as directly as we can here). Note that in @\texttt{lang}, macros can be seen as a mode of use of the term constructor \verb|Call|, as we can give the macro a singleton type (like Python modules, discussed above) that performs the desired rewriting in \verb|ana_Call| and \verb|syn_Call|.

%Generic functions are a novel strategy for {\it function polymorphism} -- defining functions that operate over more than a single type. In @\texttt{lang}, generic functions are implicitly polymorphic and can be called with arguments of {\it any type that supports the operations used by the function}. This is related to structural polymorphism \cite{malayeri2009structural}. Structural types make explicit the structure required of an argument, unlike generic functions, which are only given singleton types because the structure may depend on the semantics of an active type. 
%Structural typing can be compared to the more \emph{ad hoc} approach taken by dynamically-typed languages like Python itself, sometimes called ``duck typing''. It is also comparable to the C++ template system, as discussed previously.

{\it Operator overloading} \cite{vanWijngaarden:Mailloux:Peck:Koster:Sintzoff:Lindsey:Meertens:Fisker:acta:1975} and {\it metaobject dispatch} \cite{Kiczales91} are run-time protocols that translate operator invocations into function calls. The function is typically selected according to the dynamic tag or value of one or more operands. These protocols share the notion of {\it inversion of control} with our strategy for targeted expressions. However, our strategy is a {\it compile-time} protocol. Note that we used Python's operator overloading and metaobject protocol for convenience in the static language. %Our dispatch protocol was designed to be conceptually similar to Python's operator overloading protocol \cite{python}.

Language-external mechanisms for creating and composing dialects (e.g. extensible compilers like Xoc \cite{conf/asplos/CoxBCKK08} or language workbenches \cite{erdweg2013state}) have ambiguity problems, and so they might benefit from the type constructor oriented view that we propose here as well. We argue that if ambiguities cannot occur and safety is guaranteed, there is no reason to leave the mechanism outside the language. Though we do not support syntax extension (other than via creative reuse of string literals), we argue that this is actually a benefit: we can use a variety of existing tools and avoid many facets of the expression problem \cite{wadler1998expression} in this way.

Typed Racket and other typed LISPs also add a type system to a fixed syntax atop a dynamically typed core \cite{TypedScheme2008}. However, these treat the semantics as a ``bag of rules'', so adding new rules can cause  ambiguities. Our work (particularly @$\lambda$, with its use of lists ubiquitously) provides a blueprint for a typed LISP oriented around type constructors, rather than rulesets. Type constructors are a natural organizational unit. For example, Harper's textbook, upon which we base our terminology and abstract syntax in this paper, identifies languages directly as a collection of type constructors \cite{pfpl}.

Bidirectional type systems are increasingly being used in practical settings, e.g. in Scala and C\#. They are useful in producing good error messages (which our mechanism shows can be customized) and help avoid the need for redundant type annotations while avoiding decidability limitations associated with whole-program type inference. Bidirectional techniques have also been used for adding refinement types to languages like ML and Twelf \cite{Lovas08abidirectional}. Refinement types can add stronger static checking over a fixed type system (analagous to formal verification), but cannot be used to add new operations directly to the language. Our use of a dynamic semantics for the static language relates to the notion of \emph{type-level computation}, being explored in a number of languages (e.g. Haskell) for reasons other than extensibility. %Combining the approaches may be a fruitful avenue for future work.

Recent work we have done on the Wyvern programming language uses bidirectional typechecking to control aspects of literal syntax in a manner similar to, but somewhat more flexible than, how we treat introductory forms \cite{TSLs}. Wyvern is its own language dialect and does not have an extensible type system, supporting only desugarings like SugarJ \cite{erdweg2011sugarj}. The Wyvern formalism guarantees hygiene, using an approach that is also likely applicable in the setting of this paper.
%\begin{figure*}[t]
%\small
%\vspace{-10pt}
%{\onecolumn
%\begin{longtable}{l l@{}l c@{}c c@{}c c@{}c c@{}c}
%%\toprule%
%
%&&  &&  {\bfseries Itself} && {\bfseries Extensible} && {\bfseries Extensible} && {\bfseries Unambiguous}\\
%
%{\bfseries Approach} && {\bfseries Examples} && {\bfseries Library} && {\bfseries Syntax} && {\bfseries Type System} && {\bfseries Composition}  \\
%
%\cmidrule(l){1-1} \cmidrule(l){2-3} \cmidrule(l){4-5} \cmidrule(l){6-7} \cmidrule(l){8-9} \cmidrule(l){10-11} 
%
%\endhead
%
%
%Active Types && && \CIRCLE && \Circle && \CIRCLE && \CIRCLE  \\
%\myrowcolour%
%Type-Specific Literals && \cite{Omar:2013:TWP:2489812.2489815} && \Circle && \CIRCLE && \Circle && \CIRCLE  \\
%Desugaring && \cite{erdweg2011sugarj} && \Circle && \CIRCLE && \Circle && \Circle\\
%\myrowcolour%
%Typed LISP && Qi, \cite{TypedScheme2008} &&  \CIRCLE && \Circle && \CIRCLE && \Circle \\
%Attribute Grammars && \cite{VanWyk:2010:SEA}, \cite{conf/asplos/CoxBCKK08} && \Circle && \CIRCLE && \CIRCLE && \Circle\\
%%\myrowcolour
%%Static macros /  && \cite{ScalaMacros2013}, \cite{MorphJ2011}, && \Circle && \Circle && \Circle && \CIRCLE && \Circle \\
%%\myrowcolour
%%Metaprogramming &&  \cite{Sheard:1999:UMS} && && && && && \\
%%
%%Cross-Compilation && \cite{Rompf:2012:LMS, Delite2011} && \CIRCLE && \Circle && \Circle && \Circle && \CIRCLE \\
%%
%\myrowcolour%
%%EDSL Frameworks && ? && \CIRCLE && \CIRCLE && \CIRCLE && \Circle && \Circle \\
%\end{longtable}}
%\caption{Comparison to related approaches to language extensibility.}\label{relatedtable}
%\end{figure*}
%
%TODO: mention this http://blog.codeclimate.com/blog/2014/05/06/gradual-type-checking-for-ruby/

\section{Discussion}\label{discussion}
This work aimed to show that one can add static typechecking to a language like Python as a library, without undue syntactic overhead using techniques available in a growing number of languages: reflection, quasiquotes and a form of open sum for representing type constructors (here, Python's classes). The type system is not fixed, but flexibly extensible in a natural and direct manner, without resorting to complex encodings and, crucially, without the possibility of ambiguities arising at link-time. 
%We achieve this by combining a bidrectional type system and an elaboration semantics and delegating to static functions associated with user-defined type constructors, rather than operating in the usual syntax-directed manner. 
%Our examples were not trivial, showing a prototypic object system, a powerful string tracking system and immutable records coexisting as libraries within a language that initially had no static typechecking at all.

Although we do not have space to detail it here, we have conducted a substantial case study using (an older version of) @\texttt{lang}: an implementation of the entirety of the OpenCL type system (a variant of C99), as well as several extensions to it, as a library. Our operations translate to Python's lower-level FFI with OpenCL but guarantee type safety at the interface between languages. We have built a neurobiological circuit simulation framework atop this library as well. Based on our examples in this paper and this admittedly anecdotal experience report, we submit that active type constructors deserve broad consideration by the research community as a foundational technique for structuring typed programming languages. We are currently refactoring these projects against the (new) bidirectional semantics described here. The language core of  has already been implemented.

There remain several promising avenues for future work, many of which we mentioned throughout this paper. From a practical perspective, extensible implicit coercions (i.e. subtyping) using a mechanism similar to our ``handle sets'' mechanism for binary expressions would be useful. An extensible mechanism supporting type index polymorphism would also be of substantial utility and theoretical interest. Debuggers and other tools that rely not just on Python's syntax but also its semantics cannot be used directly. We believe that active types can be used to control debugging and other tools, and plan to explore this in the future. We have also not yet evaluated the feasibility of implementing more advanced type systems (e.g. linear, dependent or flow-dependent type systems) and those that require a more ``global'' view (e.g. security-oriented types) using our framework.

From a theoretical perspective, the next step is to introduce a static semantics for the internal language and the static language, so that we can help avoid issues of extension correctness and guarantee extension safety (by borrowing techniques from the typed compilation literature). An even more interesting goal would be to guarantee that extensions are mutually conservative: that one cannot weaken any of the guarantees of the other. We believe that by enforcing strict abstraction barriers between extensions, we can approach this goal. Bootstrapping the @\texttt{lang} compiler would be an interesting direction to explore to enable this.
% Different dialects can be difficult to combine manually, as we discussed in the introduction. We have implemented the full OpenCL programming language (a variant of C99) as a library (though the details are left for future work), showing that such languages can instead be written as libraries, rather than as dialects.
% 
%\subsection{Compositional Reasoning}\label{safety}
%Every statement and expression in an @\texttt{lang} function is governed by exactly one active type, determined by the dispatch protocol, or by the single active base associated with the @\texttt{lang} function. The representational consistency check ensures that compilation is successful without requiring that types that abstract over other types (e.g. container types) know about their internal implementation details. Together, this permits compositional reasoning about the semantics of @\texttt{lang} functions even in the presence of many extensions. This stands on contrast to many prior approaches to extensibility, where extensions could insert themselves into the semantics in conflicting ways, making the order of imports matter (see Sec. \ref{related}).

%\begin{enumerate}
%\item We address the ``chicken-and-egg'' problem by designing @\texttt{lang}, an extensible language  embedded within Python \cite{Politz:2013:PFM:2509136.2509536,python} entirely as a library. The top-level of an @\texttt{lang} file is a \emph{compilation script} written in Python and Python serves as @\texttt{lang}'s \emph{type-level language}, as we will discuss.  @\texttt{lang} and Python thus share a common syntax and package system, so users of @\texttt{lang} can directly leverage established tools, infrastructure and coding standards. Functions marked by a decorator as under the control of @\texttt{lang} are, unlike Python functions, \emph{statically} typechecked. This makes it possible to rule out many potential issues ahead of execution, and thus  avoid  costly dynamic checks. %The semantics of  define run-time behavior.
%\item  % Instead, @\texttt{lang} has a statically-typed semantics that can be extended by users from within libraries. 
%There is \emph{no particular type system built into @\texttt{lang}}. Instead, a type is any  object implementing the \lstinlinep{ace.Type} interface constructed by the compilation script. We call these \emph{active types}. When statically assigning a type to a compound form, e.g. \lstinlinep{e.x}, the @\texttt{lang} {compiler} delegates to the type of a subexpression, e.g. to the type of \lstinlinep{e}, to determine how to assign a type to the expression as a whole. As we will show, this {type-directed protocol}, \emph{active typing}, permits the expression of a rich variety of type systems as libraries. Unlike syntax-directed protocols, there cannot be ambiguities when separately defined type sytems are combined (types control non-overlapping sets of expressions by construction). Base cases, e.g.  variables and certain statement forms, are handled on a {per-function basis} by a \emph{base semantics}, also a compile-time object implementing an interface, \lstinlinep{ace.Base}.
%\end{enumerate}
%% As a result, clients are able to import any combination of extensions with the confidence that link-time ambiguities cannot occur .
%
%The dynamic behavior of an @\texttt{lang} function is determined by translation to a \emph{target language}. We begin by  targeting Python, then discuss targeting OpenCL and CUDA, lower-level languages used to program many-core processors (e.g. GPUs). Active types and bases control translation by the same type-directed protocol that governs type assignment. We call this phase \emph{active translation}. A \emph{target} mediates this process and is also a user-defined compile-time object implementing an interface, \lstinlinep{ace.Target}. 
%%One base or type can support multiple targets. 
%When targeting a typed language, care must be taken to ensure that well-typed @\texttt{lang} expressions have well-typed translations, so we integrate a technique for compositionally reasoning about compiler correctness (developed for the TIL compiler for Standard ML \cite{tarditi+:til-OLD}) into the language.
%
%%TODO
%%
%%Static type systems are powerful tools for programming language design and implementation. By tracking the type of a value statically, a typechecker can verify the absence of many kinds of errors over all inputs. This simplifies and increases the performance of the run-time system, as errors need not be detected dynamically using tag checks and other kinds of assertions.
%
%It is legitimate to ask, however, why dynamically-typed languages are so widely-used in multiple domains. Although slow and difficult to reason about, these languages generally excel at satisfying the criteria of \textbf{ease-of-use}. More specifically, Cordy identified the principle of \emph{conciseness} as elimination of
%redundancy and the availability of reasonable defaults \cite{cordy1992hints}. Statically-typed languages, particularly those that professional end-users are exposed to, are often verbose, requiring explicit and often redundant type annotations on each function and variable declaration, separate header files, explicit template headers and  instantiation and other sorts of annotations.  Dynamically-typed languages, on the other hand, avoid most of this overhead by relying on support from the run-time system. @\texttt{lang} was first conceived to explore the question: \emph{does conciseness require run-time mechanisms, or can one develop a statically-typed language with the same usability profile?}
%
%
%Rather than designing a new syntax, or modifying the syntax of an existing language, we chose to utilize, \emph{without modification}, the syntax of an existing language, Python. This choice was not arbitrary, but rather a key means by which @\texttt{lang} satisfies extrinsic design criteria related to familiarity and tool support. Researchers often dismiss the importance of syntax. By repurposing a well-developed syntax, they no longer need to worry about the ``trivial'' task of implementing it.
%
%Most programming languages are {\em monolithic} -- a collection of primitives are given first-class treatment by the language implementation, and users can only creatively combine them to implement  abstractions of their design. Although highly-expressive general-purpose mechanisms have been developed (such as object systems or functional datatypes), these may not suffice when researchers or domain experts wish to evolve aspects of the type system, exert control over the representation of data, introduce specialized run-time mechanisms, if defining an abstraction in terms of existing mechanisms is unnatural or verbose, or if custom error messages are useful. In these situations, it would be desirable to have the ability to modularly extend existing systems with new compile-time logic and be assured that such extensions will never interfere with one another when used in the same program.
%
%Python does not truly prevent extensions from interfering with one another because it lacks, e.g., data hiding mechanisms. One type could change the implementation of another by replacing its methods, for example. But these kinds of conflicts do not occur ``innocently'', as conflicts between two libraries in SugarJ that use similar syntax might.
% 
%
%

Not all extensions will be useful. Indeed, some language designers worry that offering too much flexibility to users leads to abuse. This must be balanced with the possibilities made available by a vibrant ecosystem of competing statically-typed abstractions that can be developed and deployed as libraries, and thus more easily evaluated in the wild. With an appropriate community process, this could lead to faster adoption of ideas from the research community, and quicker abandonment of mistakes. %We also anticipate that coding guidelines and tools mandating the use of abstractions that are known to have certain desirable properties will replace designer-mandated enforcement.
% 
%This paper is presented as a language design paper. The theoretical foundations of this work lie in type-level computation. We have developed a simplified, type-theoretic formalism of active typechecking and translation, where we prove the safety properties we have outlined here, as well as several additional ones that are only possible to achieve using a typed metalanguage (currently in submission at ESOP 2014). The work described here  extends that mechanism in several directions: @\texttt{lang} supports a rich syntax, makes syntax trees available to extensions, supports a pluggable backend and per-function base semantics. It is implemented in an existing language and supports a wider range of practical extensions.

%\acks
%The author is grateful to Jonathan Aldrich, Robert Bocchino and anonymous referees for their useful suggestions. This work was funded by the DOE Computational Science Graduate Fellowship under grant number DE-FG02-97ER25308.
%Acknowledgments, if needed.

% We recommend abbrvnat bibliography style.

\bibliographystyle{abbrv}

% The bibliography should be embedded for final submission.

\bibliography{../research}
%\softraggedright
%P. Q. Smith, and X. Y. Jones. ...reference text...
%
%\appendix
%\section{Source Code Emission}
%Portions of the \lstinlinep{Target} class related to source code emission are shown. If compilation occurred externally, the \lstinlinep{acec} library asks the targets used by the top-level typed functions to generate one or more files by deciding on a name and file extension on the basis of the name of the compilation script, and emitting code and \emph{snippets}, which are simply strings corresponding to functions, type declarations, imports and other top-level entities in the target language, each inserted at a \emph{location}. Each file being generated is a location, and locations might also be nested, depending on the language (e.g. the import block may be a separate location when generating Java, or the body of the \lstinlinep{<head>} tag when generating HTML+CSS+Javascript). Snippets are only added once to a particular location (so imports are not duplicated, for example).
%%We leave layering extensions to Python to created typed targets as future work.
%
%\begin{codelisting}
%\begin{lstlisting}
%class PyTarget(ace.Target):
%  class PyAST(object): """Base class for ASTs"""
%  target_type = ast
%  class Attribute(PyAST, ast.Attribute):
%    def emit(self, cg):
%      cg.append(self.value.emit(cg), '.', self.attr)
%  # ... 
%  class AnonModule(object): 
%    def __init__(self, name): 
%      self.name = name
%      self.guid = Guid()
%      
%    def emit(self, cg):
%      import_loc = cg['imports']
%      _cache = AnonModule._idx_cache[import_loc]
%      if self not in _cache:
%        anon_name = '__ace_' + str(len(_cache)+1)
%        imp_stmt = ('import ' + self.name + ' as ' + 
%          anon_name + cg.newline)
%        _cache[self] = (anon_name, imp_stmt)
%      else: 
%        anon_name, imp_stmt = _cache[self]
%      imports.add_snippet(imp_stmt)
%      cg.append(anon_name)
%      
%  def emit(self, script_name, cg):
%    file_name = '_' + script_name + '.py'
%    if file_name not in generator:
%      if 'imports' not in generator[file_name]:
%        generator[file_name].push_loc('imports')
%      if 'main' not in format_set[file_name]:
%        generator[file_name].push_loc('main')
%    translation.emit((file_name, 'main'))
%\end{lstlisting}
%\caption{Portions of the target showing how non-local code emission works.}
%\label{cg}
%\end{codelisting}
%
%\section{Kinding}
%\begin{figure}[t]
%\small
%%\vspace{-15pt}
%$\fbox{\inferrule{}{\tKindX{\tau}{\kappa}}}$
%%~~~$\chi ::= \cdot \pipe \checkmark$
%\begin{mathpar}
%\small\inferrule[k-var]{
%}{
%  \tKind{\tvarCtxX{t}{\kappa}}{\fvalCtx}{\tvar{t}}{\kappa}
%}
%~~~~
%\inferrule[k-arrow-i]{
%  \tKind{\tvarCtxX{t}{\kappa_1}}{\fvalCtx}{\tau}{\kappa_2}
%}{
%  \tKindX{\tlam{t}{\kappa_1}{\tau}}{\karrow{\kappa_1}{\kappa_2}}
%}
%~~~~
%\inferrule[k-arrow-e]{
%  \tKindX{\tau_1}{\karrow{\kappa_1}{\kappa_2}}\\\\
%  \tKindX{\tau_2}{\kappa_1}
%}{
%  \tKindX{\tapp{\tau_1}{\tau_2}}{\kappa_2}
%}
%
%\text{\color{gray} (kinding for integers, labels, lists, products and sums also standard)}
%
%%%
%%%\inferrule{ }{
%%% \tKindX{\tstr{str}}{\kstr}
%%%}(\text{str}^I_\tau)
%%%
%%%\inferrule{ }{
%%% \tKindX{\tunit}{\kunit}
%%%}(\text{1}^I_\tau)
%%%
%%%\inferrule{
%%% \tKindX{\tau_1}{\kappa_1}\\
%%% \tKindX{\tau_2}{\kappa_2}
%%%}{
%%% \tKindX{\tpair{\tau_1}{\tau_2}}{\kpair{\kappa_1}{\kappa_2}}
%%%}({\times}^I_\tau)
%%%
%%%\inferrule{
%%% \tKindX{\tau}{\kpair}{\kappa_1}{\kappa_2}
%%%}{
%%% \tKindX{\tfst{\tau}}{\kappa_1}
%%%}({\times}^{E1}_\tau)
%%%
%%%\inferrule{
%%% \tKindX{\tau}{\kpair}{\kappa_1}{\kappa_2}
%%%}{
%%% \tKindX{\tsnd{\tau}}{\kappa_2}
%%%}({\times}^{E2}_\tau)
%%%\inferrule{ }{
%%%\inferrule{
%%% \tKindX{\tau_{1}}{\kappa_{1}}\\
%%% \tKindX{\tau_{2}}{\kappa_{2}}
%%%}{
%%% \tKindX{\tpair{\tau_{1}}{\tau_{2}}}{\kpair{\kappa_{1}}{\kappa_{2}}}
%%%}~(\times_\tau)
%%%
%%%\inferrule{
%%% \tKindX{\tau}{\kpair{\kappa_{1}}{\kappa_{2}}}
%%%}{
%%% \tKindX{\tfst{\tau}}{\kappa_{1}}
%%%}~(\text{fst}_\tau)
%%%
%%%\inferrule{
%%% \tKindX{\tau}{\kpair{\kappa_{1}}{\kappa_{2}}}
%%%}{
%%% \tKindX{\tsnd{\tau}}{\kappa_{2}}
%%%}~(\text{snd}_\tau)
%%%
%%% TODO: list
%\inferrule[k-eq]{
% \kEq{\kappa}\\
% \tKindX{\tau_1}{\kappa}\\
% \tKindX{\tau_2}{\kappa}\\\\
% \tKindX{\tau_3}{\kappa'}\\
% \tKindX{\tau_4}{\kappa'}
%}{
% \tKindX{\tifeq{\tau_1}{\tau_2}{\kappa}{\tau_3}{\tau_4}}{\kappa'}
%}
%
%\inferrule[k-ty-i]{
% \fval{tycon}{\kappaidx}{-}{-} \in \fvalCtx\\\\
% \tKind{\tvarCtx}{\fvalCtx}{\tauidx}{\kappaidx}
%}{
% \tKindX{\ttype{tycon}{\tauidx}}{\kTypeBlur}
%}
%
%\inferrule[k-ty-e]{
% \tKindX{\tau}{\kTypeBlur}\\
% \fval{tycon}{\kappaidx}{-}{-} \in \fvalCtx\\\\
% \tKind{\tvarCtxX{x}{\kappaidx}}{\fvalCtx}{\tau_1}{\kappa}\\
% \tKindX{\tau_2}{\kappa}
%}{
% \tKind{\tvarCtx}{\fvalCtx}{\tfamcase{\tau}{tycon}{x}{\tau_1}{\tau_2}}{\kappa}
%}
%~~~~~~
%\inferrule[k-d-i]{
% \tKindX{\tau_1}{\kTypeBlur}\\\\
% \tKindX{\tau_2}{\kITerm}
%}{
% \tKindX{\tden{\iota}{\tau}}{\kDen}
%}
%
%%\inferrule[k-d-i-checked]{
%%  \tKind{\Delta}{\fvalCtx}{\tau}{\kTypeBlur}\\\\
%%  \isItermX{\ibar}
%%}{
%%  \tKind{\Delta}{\fvalCtx}{\tden{\ibar}{\tau}^\checkmark}{\kDen}
%%}
%%~~~~~~~
%\inferrule[k-d-e]{
% \tKindX{\tau}{\kDen}
%}{
% \tKindX{\ttypeof{\tau}}{\kTypeBlur}
%}
%%
%%\inferrule[iterm-intro]{
%%  \isIterm{\tvarCtx}{\emptyctx}{\fvalCtx}{\iota}
%%}{
%%  \tKindX{\titerm{\iota}}{\kITerm}
%%}
%
%\inferrule[k-reptype-i]{
% \isItype{\tvarCtx}{\fvalCtx}{\sbar}
%}{
% \tKindX{\titype{\sigma}}{\kIType}
%}
%\end{mathpar}
%%$\fbox{$\isItermX{\ibar}$}$
%%\begin{mathpar}
%%\inferrule[k-i-var]{ }{
%%  \isItermX{\evar{x}}
%%}
%%~~~~~~
%%\inferrule[k-i-fix]{
%%  \isItypeX{\sbar}\\
%%  \isItermX{\ibar}
%%}{
%%  \isItermX{\ifix{\evar{x}}{\sbar}{\ibar}}
%%}
%%~~~~~~
%%\inferrule[k-i-transof]{
%%  \tKindX{\tau}{\kDen}
%%}{
%%  \isItermX{\itransof{\tau}}
%%}
%%%\inferrule[i-lam-kinding]{
%%% \isItypeX{\sbar}\\\\
%%% \isItermX{\ibar}
%%%}{
%%% \isItermX{\ilam{\evar{x}}{\sbar}{\ibar}}
%%%}
%%
%%\text{\color{gray} (omitted rules are trivially recursive, like \textsc{k-i-fix})}
%%%\inferrule[iterm-dereify-kinding]{
%%% \tKindX{\tau}{\kITerm}
%%%}{
%%% \isItermX{\iup{\tau}}
%%%}
%%%
%%\end{mathpar}
%$\fbox{$\isItypeX{\sbar}$}$
%\begin{mathpar}
%%\inferrule[i-int-kinding]{ }{
%%  \isItypeX{\dint}
%%}
%%
%%\inferrule[i-prod-kinding]{
%%  \isItypeX{\sigma_1}\\\\
%%  \isItypeX{\sigma_2}
%%}{
%%  \isItypeX{\dpair{\sigma_1}{\sigma_2}}
%%}
%%
%\inferrule[k-s-unquote]{
% \tKindX{\tau}{\kIType}
%}{
% \isItypeX{\dup{\tau}}
%}
%
%%\inferrule[k-s-repof]{
%%  \tKindX{\tau}{\kTypeBlur}
%%}{
%%  \isItypeX{\trepof{\tau}}
%%}
%%
%\text{\color{gray} (omitted forms are trivially recursive)}
%\end{mathpar}
%\caption{\small Kinding for type-level terms. The kinding context $\tvarCtx$ maps type variables to kinds.}
%\label{tlkind}
%\end{figure} 
%Some of the kinding rules for type-level terms are shown.
%
%\section{Complex Numbers}
%We have implemented complex numbers (which have a logic similar to decimals, which we added late in the paper cycle). The API is somewhat dated, apologies for
%
%If \lstinlinep{c} is a complex number, then \lstinlinep{c.ni} and \lstinlinep{c.i} are its non-imaginary and imaginary components, respectively. These expressions are of the form \lstinlinep{Attribute}, so the typechecker calls \lstinlinep{type_Attribute} (line \ref{cplx}.7).
%This method receives the compilation context, \lstinlinep{context}, and the abstract syntax tree of the expression, \lstinlinep{node} and must return a type assignment for the node, or raise an \lstinlinep{ace.TypeError} if there is an error. In this case, a type assignment is possible if the attribute name is either \lstinlinep{"ni"} or \lstinlinep{"i"}, and an error is raised otherwise (lines \ref{cplx}.8-\ref{cplx}.10). We note that error messages are an important and sometimes overlooked facet of {ease-of-use} \cite{marceau2011measuring}. A common frustration with using general-purpose abstraction mechanisms to encode an abstraction is that they can produce  verbose and cryptic error messages that reflect the implementation details instead of the semantics. @\texttt{lang} supports custom error messages.%Active types permit direct encoding of semantics and customization of error messages.
%
%Complex numbers also support binary arithmetic operations partnered with both other complex numbers and with non-complex numbers, treating them as if their imaginary component is zero. The typechecking rules for this logic is implemented on lines \ref{cplx}.17-\ref{cplx}.29. Arithmetic operations are usually symmetric, so the dispatch protocol checks the types of both subexpressions for support. To ensure that the semantics remain deterministic in the case that both types support the binary operation, @\texttt{lang} asks the left first (via \lstinlinep{type_BinOp_left}), asking the right (via \lstinlinep{type_BinOp_right}) only if the left indicates an error. In either position, our implementation begins by recursively assigning a type to the other operand in the current context via the \lstinlinep{context.type} method (line \ref{cplx}.24). If supported, it applies the C99 rules for arithmetic operations to determine the resulting type (via \lstinlinep{c99_binop_t}, not shown). 
%
%Finally, a complex number can be constructed inside an @\texttt{lang} function using @\texttt{lang}'s special constructor form: \lstinlinep{[clx.Cplx](3,4)} represents $3+4i$, for example. The term within the braces is evaluated at \emph{compile-time}. Because \lstinlinep{clx.Cplx} evaluates not to an active type, but to a class, this form is assigned a type by handing control to the class object via the \emph{class method} \lstinlinep{type_New}. It operates as expected, extracting the types of the two arguments to construct an appropriate complex number type (lines \ref{cplx}.50-\ref{cplx}.57), raising a type error if the arguments cannot be promoted to a common type according to the rules of C99 or if two arguments were not provided.% (an exercise for the reader: modify this method to also allow a single argument for when the imaginary part is 0). 
%
%As seen in Listing \ref{mapout}, we are implementing complex numbers internally using OpenCL vector types, like \lstinlinep{int2}. Let us look first at \lstinlinep{trans_New} on lines \ref{cplx}.54-\ref{cplx}.60, where new complex numbers are translated to vector literals by invoking \lstinlinep{target.VecLit}. This will ultimately  generate the necessary OpenCL code, as a string, to complete compilation (these strings are not directly manipulated by extensions, however, to avoid problems with, e.g. precedence). For it to be possible to reason compositionally about the correctness of compilation, all complex numbers must translate to terms in the target language that have a consistent target type. The \lstinlinep{trans_type} method of the \lstinlinep{ace.ActiveType} associates a type in the target language, here a vector type like \lstinlinep{int2}, with the active type. @\texttt{lang} supports a mode where this \emph{representational consistency} is dynamically checked during compilation (requiring that the active target know how to assign types to terms in the target language, which can be done  for our OpenCL target as of this writing).
%
%The translation methods for attributes (line \ref{cplx}.12) and binary operations  (line \ref{cplx}.31) proceed in a straightforward manner. The context provides a method, \lstinlinep{trans}, for recursively determining the translation of subexpressions as needed. Of note is that the translation methods can assume that typechecking succeeded. For example, the implementation of \lstinlinep{trans_Attribute} assumes that if \lstinlinep{node.attr} is not \lstinlinep{'ni'} then it must have been \lstinlinep{'i'} on line \ref{cplx}.14, consistent with the implementation of \lstinlinep{type_Attribute} above it. Typechecking and translation are separated into two methods to emphasize that typechecking is not target-dependent, and to allow for more advanced uses, like type refinements and hypothetical typing judgements, that we do not describe here.
%
%\begin{codelisting}
%\lstinputlisting{listing9.py}
%\caption{\texttt{[}in \texttt{examples/clx.py]} The active type family \texttt{Ptr} implements the semantics of OpenCL pointer types.}
%\label{cplx}
%\end{codelisting}
%\section{Expressiveness}\label{examples}
%Thus far, we have focused mainly on the OpenCL target and shown examples of fairly low-level active types: those that implement OpenCL's primitives (e.g. \lstinlinep{clx.Ptr}), extend them in simple but convenient ways (e.g. \lstinlinep{clx.Cplx}) and those that make interactive execution across language boundaries safe and convenient (e.g. \lstinlinep{clx.NPArray}). 
%@\texttt{lang} was first conceived to answer the question: \emph{can we build a statically-typed language with the semantics of a C but the syntax and ease-of-use of a Python?} We submit that the work described above has answered this question in the affirmative. 
%
%But @\texttt{lang} has proven useful for more than low-level tasks like programming a GPU with OpenCL. We now describe several interesting extensions that implement the semantics of primitives drawn from a range of different language paradigms, to justify our claim that these mechanisms are highly expressive. 
%
%\subsection{Growing a Statically-Typed Python Inside an @\texttt{lang}}
%@\texttt{lang} comes with a target, base and type implementing Python itself: \lstinlinep{ace.python.python}, \lstinlinep{ace.python.base} and \lstinlinep{ace.python.dyn}. These can be supplemented by additional active types and used as the foundation for writing actively-typed Python functions. These functions can either be compiled ahead-of-time to an untyped Python file for execution, or be immediately executed with just-in-time compilation, just like the OpenCL examples we have shown. Many of the examples in the next section support this target in addition to the OpenCL target we have focused on thus far.
%
%\section{Recursive Labeled Sums}
%Listing 12 shows how recursive labeled sums (i.e. functional datatypes) would look.
%\begin{codelisting}
%\lstinputlisting{datatypes_t.py}
%\caption{\texttt{[datatypes\_t.py]} An example using statically-typed functional datatypes.}
%\label{datatypest}
%\end{codelisting}
%\begin{codelisting}
%\lstinputlisting{datatypes.py}
%\caption{\texttt{[datatypes.py]} The dynamically-typed Python code generated by running \texttt{acec datatypes\_t.py}.}
%\label{datatypes}
%\end{codelisting}
%Listing \ref{datatypest} shows an example of the use of statically-typed functional datatypes (a.k.a. recursive labeled sum types) together with the Python implementation just described. It shows two syntactic conveniences that were not mentioned previously: 1) if no target is provided to \lstinlinep{ace.fn}, the base can provide a default target (here, \lstinlinep{ace.python.python}); 2) a concrete function can be generated immediately by providing a type assignment after \lstinlinep{ace.fn} using braces. Listing \ref{datatypest} generates Listing \ref{datatypes}.
%
%Lines \ref{datatypest}.3-\ref{datatypest}.11 define a function that generates a recursive algebraic datatype representing a tree given a name and another @\texttt{lang} type for the data at the leaves. This type implemented by the active type family \lstinlinep{fp.Datatype}. A name for the datatype and the case names and types are provided programmatically on lines \ref{datatypest}.3-\ref{datatypest}.8. To support recursive datatypes, the case names are enclosed within a lambda term that will be passed a reference to the datatype itself. These lines also show two more active type families: units (the type containing just one value), and immutable pairs. Line \ref{datatypest}.13 calls this function with the @\texttt{lang} type for dynamic Python values, \lstinlinep{py.dyn}, to generate a type, aliased \lstinlinep{DT}. This type is implemented using class inheritance when the target is Python, as seen on lines \ref{datatypes}.1-\ref{datatypes}.13. For C-like targets, a \lstinlinep{union} type can be used (not shown).
%
%The generic function \lstinlinep{depth_gt_2} demonstrates two features. First, the base has been setup to treat the final expression in a top-level branch as the return value of the function, consistent with typical functional languages. Second, the \lstinlinep{case} ``method'' on line \ref{datatypest}.17 creatively reuses the syntax for dictionary literals to express a nested pattern matching construct. The patterns to the left of the colons are not treated as expressions (that is, a type and translation is never recursively assigned to them). Instead, the active type implements the standard algorithm for ensuring that the cases cover all possibilities and are not redundant \cite{pfpl}. If the final case were omitted, for example, the algorithm would statically indicate an error, just as in statically-typed functional languages like ML.
%
%%\subsection{Nested Data Parallelism}
%%Functional constructs have shown promise as the basis for a nested data-parallel parallel programming model. Many powerful parallel algorithms can be specified by composing primitives like \lstinlinep{map} and \lstinlinep{reduce} on persistent data structures like lists, trees and records. Data dependencies are directly encoded in these specifications, and automatic code generation techniques have shown promise in using this information to automatically execute these specifications on concurrent hardware and networks. The Copperhead programming language, for example, is based in Python as well and allows users to express functional programs over lists using common functional primitives, compiling them to CUDA code \cite{catanzaro2011copperhead}. By combining active types and the metaprogramming facilities described in Sec. 2 to implement optimizations, like fusion, on the typed syntax trees available from concrete functions, these same techniques can be implemented as an @\texttt{lang} library as well.
%
%\subsection{Product Types}
%Product types (types that group heterogeneous values together) like structs, unions, tuples, records and objects can all be implemented as @\texttt{lang} type families  parameterized by a dictionary mapping field names (indices, in the case of tuples) to types. Each family differs slightly in the semantics of the operations available. Structs support mutation, for example, while records do not. Object systems typically introduce additional logic for calling methods, binding special variables like \lstinlinep{this}, and accessing information through an inheritance hierarchy. Listing \ref{ooclxfppy} shows an example of each of these. The prototypic object system delegates to a backing record if a field is not available in the foreground. This example also demonstrates cross-compilation to C99, supported for a subset of operations in \lstinlinep{clx}. All three abstractions are implemented as simple \lstinlinep{struct}s, as can be seen in Listing \ref{ooclxfpc}. This example further demonstrates the use of argument annotations to generate concrete functions (line \ref{ooclxfppy}.9), which is a feature only available when using Python 3.3+. 
%
%\subsection{Distributed Programming}
%Many distributed programming abstractions can be understood as implementing object models with complex forms of dynamic dispatch. For example, in Charm++, dynamic dispatch involves message passing over a dynamically load-balanced network \cite{kale2009charm++}. While Charm++ is a sophisticated system, and we do not anticipate that implementing it as an @\texttt{lang} extension would be easy, it is possible to do so by targeting a system like Adaptive MPI, which exposes the Charm++ runtime system \cite{kale2009charm++}. %Erlang and other message-passing and actor-based systems can also be considered in this manner -- processes can be thought of as objects, channels as methods and messages as calls.
%
%%\subparagraph{Annotation and Extension Inference}
%%In addition to type annotations, OpenCL normally asks for additional annotations in a number of other situations.  Users can annotate functions that meet certain requirements  with the \lstinlinep{__kernel} attribute, indicating that they are callable from the host. The \lstinlinep{OpenCL} backend can check these and add this annotation automatically. Several types (notably, \lstinlinep{double}) and specialized primitive functions require that an OpenCL extension be enabled via a \lstinlinep{#pragma}. The OpenCL backend automatically detects many of these cases as well, adding the appropriate \lstinlinep{#pragma} declaration. An example of this can be seen in Listing \ref{mapout}, where the use of the \lstinlinep{double} type triggers the insertion of the \lstinlinep{cl_khr_f64} extension.
%%
%
%A number of recent languages designed for distributed programming on clusters use a partitioned global address space model, e.g. Chapel \cite{chapel}. These languages provide first-class support for accessing data transparently across a massively parallel cluster. Their type systems track information about \emph{data locality} so that the compiler can emit more efficient code. The extension mechanism can also track this information in a manner analogous to how address space information is tracked in the OpenCL example above, and target an existing portable run-time such as GASNet \cite{bonachea2002gasnet}.
%
%\subsubsection{Units of Measure}
%A number of domain-specific type systems can be implemented within @\texttt{lang} as well. For example, prior work has considered tracking units of measure (e.g. grams) statically to validate scientific code \cite{conf/cefp/Kennedy09}. This cannot easily be implemented using many existing abstraction mechanisms because this information should only be maintained statically to avoid excessive run-time overhead associated with tagging/boxing, and the typechecking logic is reasonably complex. The @\texttt{lang} extension mechanism allows this information to be tracked in the type system, but not included during translation, by a mechanism similar to how address space information is tracked with pointers. Because @\texttt{lang} extensions can use Python, the needed logic can be implemented directly.
%
%
%\subsection{OOFPCLX}
%\begin{codelisting}[t]
%\lstinputlisting{ooclxfp.py}
%\caption{\texttt{[ooclxfp.py]} An example combining structs and immutable records using a prototype-based object system, cross-compiled to C99. Uses Python 3 argument annotations.}
%\label{ooclxfppy}
%\end{codelisting}
%\begin{codelisting}[t]
%\lstinputlisting[style=OpenCL]{ooclxfp.c}
%\caption{\texttt{[ooclxfp.c]} The C99 code generated by running \texttt{acec ooclxfp.py}.}
%\label{ooclxfpc}
%\end{codelisting}

\end{document}

### 2014-06-04 atlang-gpce14/atlang-gpce14.tex
atlang-gpce14 submission
static in subtitle

@@
 \title{Active Type Constructors}
-\subtitle{Unambiguously Growing a Type System Inside a Python}
+\subtitle{Unambiguously Growing a Static Type System Inside a Python}
 %\titlebanner{{\tt \textcolor{Red}{{\small Under Review -- distribute within CMU only.}}}}        % These are ignored unless
@@
 \end{lstlisting}
-\caption{Execution never proceeds into a function with a type error.}
+\caption{Execution never proceeds into a function with a type error when using @$\texttt{lang}$ for implicit compilation.}
 \label{oops-exec}

### 2014-07-09 att-popl15/att-popl15.tex
POPL15 submission

\documentclass[9pt,preprint]{sigplanconf}
\usepackage{amsmath,tabu}
\usepackage{cancel}
\usepackage{amssymb} 
\usepackage{amsthm}
\usepackage{graphicx}
\usepackage[usenames,dvipsnames,svgnames,table]{xcolor}
\newcommand{\copyleft}{\reflectbox{\sffamily\copyright}}
\usepackage{ stmaryrd }
\usepackage{mathpartir}
%\usepackage{cite}
%\renewcommand{\citepunct}{,\,} % IEEEtran wants to use ],\,[ for this but that looks dumb...
\input{macros-atlam}
\input{macros-catlam}
%\usepackage{sfmath}
\renewcommand{\ttdefault}{txtt}
\usepackage{alltt}
\usepackage{listings}
\lstset{language=ML,
showstringspaces=false,
basicstyle=\ttfamily\footnotesize,
morekeywords={newcase,extends}}

\usepackage{float}
\floatstyle{ruled}
\newfloat{codelisting}{tp}{lop}
\floatname{codelisting}{Listing}

\addtolength{\floatsep}{-8px}

\newcommand{\compresslist}{
  \vspace{-1em}
  \setlength{\itemsep}{1pt}
  \setlength{\parskip}{0pt}
  \setlength{\parsep}{0pt}
}

\usepackage{url}
% url.sty was written by Donald Arseneau. It provides better support for
% handling and breaking URLs. url.sty is already installed on most LaTeX
% systems. The latest version can be obtained at:
% http://www.ctan.org/tex-archive/macros/latex/contrib/misc/
% Read the url.sty source comments for usage information. Basically,
% \url{my_url_here}.

\usepackage{placeins}

%\lefthyphenmin=4
\hyphenation{op-tical net-works semi-conduc-tor}

\usepackage{todonotes}

\lefthyphenmin=10
\sloppy

\newcommand{\F}[1]{{\sf #1}~}
\newcommand{\FF}[1]{{\sf #1}}
%\usepackage{mathptmx}
%\usepackage[T1]{fontenc}

\usepackage{microtype}
    \renewcommand{\topfraction}{0.9}  % max fraction of floats at top
    \renewcommand{\bottomfraction}{0.9} % max fraction of floats at bottom
\newcommand{\rulename}[1]{({\textsc{\textsl{#1}}})}
\def \TirNameStyle #1{\small\rulename{#1}}
\renewcommand{\MathparLineskip}{\lineskiplimit=.6\baselineskip\lineskip=.6\baselineskip plus .2\baselineskip}

\newcommand{\keyrule}{{\color{red}~$\star$}}
\newcommand{\minput}{}
\newcommand{\moutput}{^{\color{gray}{+}}}

\newtheorem{theorem}{Theorem}
\newtheorem{lemma}{Lemma}
\newtheorem{prop}{Proposition}

\begin{document}

\conferenceinfo{-}{-} 
\copyrightyear{-} 
\copyrightdata{[to be supplied]}

%\titlebanner{{\tt \textcolor{Red}{{\small Under Review -- distribute within CMU only.}}}}        % These are ignored unless
%\preprintfooter{Distribute within CMU only.}   % 'preprint' option specified.

\title{Modular Type Constructors}
\subtitle{Type System Fragments as Safely Composable Libraries}
%\subtitle{Conservatively Extending a Type System From Within}
\authorinfo{}{}{}
%\authorinfo{Cyrus Omar\and Nathan Fulton\and Jonathan Aldrich}
 %         {Carnegie Mellon University}
  %        {\{comar, nfulton, aldrich\}@cs.cmu.edu}

\maketitle
\begin{abstract} 
Abstraction providers sometimes need to directly introduce new types and operators into existing languages. 
Ideally, such typed language fragments could be separately defined and safely composed. 
Unfortunately, existing tools and techniques do not come equipped with modular reasoning principles comparable to those available for library-based embeddings packaged using a modern module system. Each combination of fragments is semantically its own dialect of the language and metatheoretic and compiler correctness results must be established monolithically. 
Recent work has started to address this problem, e.g. by showing how to modularly reason about concrete syntax. Our focus here is on safely composing separately defined type system fragments.  We organize each fragment around a type constructor, as is usual practice when describing type systems, and sidestep the difficulties of abstract syntax by statically delegating semantic control over a small, fixed abstract syntax in a type-directed manner to static logic associated with a relevant type constructor.
The paper is organized around a minimal calculus, @$\lambda$ that permits surprisingly practical examples. We establish several strong semantic guarantees, notably \emph{type safety}, \emph{stability of typing} under extension and \emph{conservativity}: 
that the \emph{type invariants} that a finite set of fragments maintain are conserved under  extension. 
This involves lifting typed compilation techniques into the semantics and enforcing an abstraction barrier around each fragment using a form of ``internal'' type abstraction. 
Conservativity then follows from classical parametricity results, so these \emph{modular type constructors} can be reasoned about like modules. %We discuss how to optionally mechanize this reasoning in Coq.%We call embeddings constructed using this strengthened formulation of active type constructors \emph{active embeddings}. A strength of the technique is that it does not require mechanized specifications and proofs to make the guarantees above, but these can be constructed, using existing proof assistants, to increase confidence that an active type constructor definition adequately implements a specified fragment.%We introduce @$\lambda$, a simply-typed lambda calculus with simply-kinded type-level computation where new indexed type and operator constructors can be declared  from within. %It is structured in four layers: at the top-level, extension providers declare new indexed type constructors and term-level operator constructors. 
%Type-level functions associated with operator constructors define their static and dynamic semantics, the latter by translation to a fixed typed nternal language. By lifting compiler correctness techniques into the language, the ``actively typed semantics'' guarantees type safety. Going further, the semantics enforce an abstraction barrier at extension boundaries that ensures that extensions are mutually conservative (i.e. they do not weaken or interfere with one another, so that they can  be used together in any combination). 
%%Totality of the type-level language guarantees decidability of type checking and equality constraints on type indices makes type equivalence decidable. 
%We intend @$\lambda$ as a minimal foundation for future work on safe language-integrated extension mechanisms for typed programming languages, but it is already quite expressive. We demonstrate by showing how a conventional concrete syntax can be introduced by type-directed dispatch to Core @$\lambda$, then discuss a number of typed language fragments that can be \emph{actively embedded} by this mechanism as orthogonal ``libraries''. %(and discuss the sorts of fragments that, as yet, cannot).
\end{abstract}

%\category{D.3.2}{Programming Languages}{Language Classifications}[Extensible Languages]
%\category{D.3.4}{Programming Languages}{Processors}[Compilers]
%\category{F.3.1}{Logics \& Meanings of Programs}{Specifying and Verifying and Reasoning about Programs}[Specification Techniques]
%\keywords
%type-level computation, typed compilation
\section{Introduction}
Typed programming languages are often described in \emph{fragments}, each defining contributions to a language's concrete syntax, abstract syntax, static semantics and dynamic semantics. 
In his textbook, Harper organizes fragments around type constructors, each introduced in a different chapter \cite{pfpl}. A language is then identified by a set of type constructors, e.g. $\mathcal{L}\{{\rightharpoonup}\,{\forall}\,{\mu}\,{\keyw{1}}\,{\times}\,{+}\}$ is the language that builds in partial function types,  polymorphic types, recursive types, nullary and binary product types and binary sum types (its syntax is shown in Figure \ref{syntax-IL}, discussed further below).
It is left implied that the metatheory developed separately is conserved when fragments like these are composed to form a language.

Luckily, fragment composition is not an everyday programming task, because fragments like these are ``general purpose'' in that they make it possible to construct \emph{isomorphic embeddings} of many other fragments as ``libraries''. For example, lists can be placed in isomorphism with polymorphic recursive sum of products, $\iforall{\alpha}{\imu{t}{\iunit + (\alpha \times t))}}$. Languages providing datatypes in the style of ML  are perhaps most directly oriented around embeddings like this (preserving type disequality requires adding some form of generativity, as ML datatypes also expose). % Leaving the core language simple makes it easier to establish its metatheory and verify that tools, like compilers, are implemented correctly.% They also have encouraging connections to logic. %For example, we do not need to define the type constructor $\fvar{list}$ (indexed by a type) as a fragment (though it is possible to do so) because there is a user-defined datatype constructor, $\texttt{list}$, parameterized by a type, that is isomorphic. If we added the $\fvar{list}$ fragment to the language, it would be entirely redundant semantically: every well-typed term of a type constructed by $\fvar{list}$ corresponds to a well-typed term and a polymorphic recursive sum type such th%Although strings and numbers can be embedded as recursive sum types, it is recognized that this is impractical, so these are also usually included as primitives.
%Establishing an isomorphic embedding of a desirable fragment in terms of general-purpose fragments is not always possible, nor are such embeddings always practical.

Unfortunately, situations do sometimes arise where using these fragments to establish an isomorphic embedding that preserves a desirable fragment's  static and dynamic semantics (including bounds specified by its cost semantics) is not possible. 
Embeddings can also sometimes be unsatisfyingly \emph{complex}, as measured by the cost of the extralinguistic computations that are needed to map in and out of the embedding and, if these  must be performed mentally, considering  cognitive metrics like  error message comprehensibility. %We will discuss specific examples below.

When an embedding is too complex, abstraction providers have a few options, discussed in Sec. \ref{prior-work}. When an embedding is not possible, however, or when these options are  insufficient, providers are often compelled to introduce these fragments by extending an existing language, thereby forming a new \emph{dialect}. To save effort, they may do so by forking existing artifacts or leverage tools like compiler generators or language frameworks, also reviewed in Sec. \ref{prior-work}. %Indeed, the proliferation of language dialects constructed for this reason might be taken as an evidence that the core language is not as ``general'', when considered comprehensively, as might be hoped. 
Within the ML lineage, dialects that go  beyond $\mathcal{L}\{{\rightharpoonup}\,{\forall}\,{\mu}\,{\keyw{1}}\,{\times}\,{+}\}$ abound:
%Reynolds, in a remark that recalls the ``Turing tarpit'' of Perlis \cite{Perl82a}, summarizes the issue \cite{Reynolds94anintroduction}: 
%\begin{quote}
%To say that any reasonable function can be expressed by some program is not to say that it can be expressed by the most reasonable program. It is clear that the language requires a novel programming style. Moreover, it is likely that certain important functions cannot be expressed by their most efficient algorithms.
%\end{quote}
%

\begin{enumerate}
\compresslist
\vspace{5px}
\item 
\textbf{General Purpose Fragments:} 
A number of variations on product types, for example, have been introduced in dialects: 
$n$-ary tuples, 
labeled tuples, 
records (identified up to reordering), 
structurally typed or row polymorphic records \cite{Cardelli:1984:SMI:1096.1098}, 
records with update and extension operators \cite{ocaml-manual}, 
mutable fields \cite{ocaml-manual}, 
%field delegation \cite{atlang-gpce14} \todo{gpce submission} 
and 
``methods'' (i.e. pure objects \cite{conf/oopsla/Aldrich13, TSLs}).\footnote{The Haskell wiki notes that ''No, extensible records are not implemented in GHC. The problem is that the record design space is large, and seems to lack local optima. And all reasonable variants break backward compatibility. As a result, nothing much happens.'' \cite{GHCFAQ}} 
 Sum types are also exposed in various ways: 
finite datatypes, 
open datatypes \cite{conf/ppdp/LohH06}, 
hierarchically open datatypes \cite{journals/toplas/MillsteinBC04}, 
polymorphic variants \cite{ocaml-manual} and 
ML-style exception types. Other generally useful fragments are also built in, e.g. \verb|sprintf| in the OCaml dialect statically distinguishes format strings from strings.

\item
\textbf{Specialized Fragments:} Fragments that track specialized static invariants to provide stronger correctness guarantees, manage unwieldy lower-level abstractions or control cost are also often introduced in dialects, e.g. for data parallelism  \cite{chakravarty2007data}, distributed programming \cite{Murphy:2007:TDP:1793574.1793585}, reactive programming \cite{mandel2005reactiveml}, authenticated data structures \cite{Miller:2014:ADS:2535838.2535851}, databases \cite{Ohori:2011:MSM:2034773.2034815} and units of measure \cite{conf/cefp/Kennedy09}.% All of these are implemented as dialects of existing languages, presumably because a strong encoding was not feasible.

\item
\textbf{Foreign Fragments:} A safe and natural foreign function interface (FFI) can be a valuable feature (particularly given this proliferation of dialects). However, this requires enforcing the type system of the foreign language in the calling language. %Using  a FFI that does not do this can lead to safety issues, even when both languages are separately known to be safe. 
%Safe FFIs generally require direct extensions to the language. 
For example, MLj builds in a safe FFI to Java \cite{Benton:1999:IWW:317636.317791}.
\end{enumerate}
\vspace{-5px}

This dialect-oriented state of affairs is, we argue, unsatisfying: a programmer can choose either a dialect supporting a principled approach to distributed programming, or one that builds in support for statically reasoning about units of measure, but there may not be an available dialect supporting both. Combining dialects is non-trivial in general, as we will discuss. Using different dialects separately for different components of a program is also untenable: components written in different dialects cannot always interface safely with one another (i.e. an FFI, item 3  above, is needed).

These problems do not arise when a fragment can be exposed as an isomorphic embedding because modern \emph{module systems} can enforce abstraction barriers that ensure that the isomorphism need only be established in the ``closed world'' of the module. This is useful because it does not impose proof obligations on clients in the ``open world''.  %Mechanisms that can help decrease the complexity of an embedding without violating abstraction barriers are thus valuable, and we will lead into our work in Sec. \ref{prior-work} by summarizing them. 
For situations where an embedding is not evident, however, mechanisms are needed that make specifying, implementing and reasoning about direct fragment composition similarly modular. Such mechanisms could ultimately be integrated directly into a language, blurring the distinction between fragments and libraries and decreasing the need for new dialects. Importing fragments that introduce new syntax and semantics   would be as easy as importing a new module is today. In the limit, the community could rely on modularly mechanized metatheory and compiler correctness results, rather than requiring heroic efforts on the part of individual research groups as the situation exists today. %Recent work has shown progress on modularly introducing new concrete syntax, reviewed in Sec. \ref{desugaring}. Our focus is on the problem of introducing new semantics.

\paragraph{Contributions} In this paper, we take substantial steps towards this goal by  constructing a minimal but powerful core calculus, @$\lambda$ (the ``actively typed'' lambda calculus). %Despite its minimality, it can host a variety of practical semantic extensions like those described above, while maintaining strong metatheoretic guarantees and, crucially, providing modular reasoning properties. %a calculus introduced briefly in recent work on \emph{active type constructors}  \cite{atlang-gpce14}\todo{TR? Arxiv?}, reviewed in Sec. \ref{overview}. 
This calculus is structured in a manner similar to the Harper-Stone semantics for Standard ML \cite{Harper00atype-theoretic}, consisting of an \emph{external language} (EL) governed by a {typed translation semantics} targeting a fixed \emph{internal language} (IL). 
Rather than building in a monolithic set of type constructors, however, the typechecking judgement is indexed by a \emph{tycon context}. Each tycon defines the semantics of its associated operators via functions written in a \emph{static language} (SL). Types are values in the SL.%, which provides a form of type-level computation. %The authors demonstrate the expressive power of this technique primarily with an implementation and a number of examples of fragments as libraries, and their core calculus .% (unlike systems that treat the type system as a ``bag of rules'', where non-determinism can arise).

We introduce @$\lambda$ by example in Sec. \ref{overview}, demonstrating its surprising expressive power and detailing its semantics, then examine its key metatheoretic properties in Sec. \ref{calculus}, beginning with \emph{type safety} and \emph{unicity of typing} and touching on  \emph{decidability of typechecking}. We then consider properties that relate to composition of tycon definitions: \emph{hygiene}, \emph{stability of typing} and a key modularity result, which we refer to as  \emph{conservativity}: any invariants that can be established about all values of a type under \emph{some} tycon context (i.e. in some  ``closed world'') are conserved in any further extended tycon context (i.e. in the ``open world'').

We combine several interesting type theoretic techniques, applying them to novel ends: 1)  a bidirectional type system permits flexible reuse of a fixed syntax; 2) the SL serves both as an extension language and as the type-level language; we give it its own statics (i.e. a \emph{kind system}); 3) we use a typed intermediate language and leverage corresponding \emph{typed compilation} techniques, here lifted into the semantics of the EL; 4) we leverage internal type abstraction implicitly as an effect during normalization of the SL to enforce abstraction barriers between type constructors. 
As a result, conservativity follows from the same elegant parametricity results that underly  abstraction theorems for module systems. 
Like modules, reasoning about these \emph{modular type constructors} does not require  mechanized specifications or proofs: correctness issues in the type constructor logic necessarily causes typechecking to fail, so even extensions that are not proven correct can be distributed and ``tested'' in the wild without compromising the integrity of an entire program (at worst, only values of types constructed by the tycon being tested may exhibit undesirable properties). 
% But mechanized proofs can be provided to verify that such failures (which do not indicate mistakes in client code but are more analagous to ``translation validation failures'' arising from the compiler \cite{Pnueli-Siegel-Singerman98}, will not occur during typechecking), 
% and that a type constructor adequately implements a separately specified fragment. We take the first steps towards such a \emph{modularly mechanized metatheory} by briefly discussing how to construct contextual embeddings of our semantics into Coq (Sec. \ref{coq}). 
%Though surprisingly expressive, we discuss some limitations of the core calculus, and suggest variants  that increase its expressive power while maintaining these key guarantees (Sec. \ref{variants}). 
We conclude by discussing some limitations, proposing some richer variants of the calculus and discussing related work (Sec. \ref{prior-work}). %Note that readers unfamiliar with prior approaches may wish to read this section first. %so that even fragments like those enumerated above can be defined as safely composable libraries, rather than in dialects that must be considered monolithically. %In the limit, building safe FFIs between every pair languages in a project is more effort than attempting to combine the languages. 
%Libraries are preferable to dialects, so there has been considerable interest in mechanisms that can make it possible to implement and reason about fragments like those above orthogonally and compose them automatically.% Mechanisms that guarantee composition will \emph{always} be safe can be integrated directly into a language as a general-purpose abstraction mechanism.  %that can fully automate the process of combining dialects defined atop a common framework (\emph{language-external mechanisms}).

\section{@$\lambda$, By Example}\label{overview}

\begin{figure}
\small
\hspace{-5px}\input{syntax-table-IL}
\caption{Syntax of {$\mathcal{L}\{{\rightharpoonup}\,{\forall}\,{\mu}\,{\iunit}\,{\times}\,{+}\}$}, our internal language (IL). We write internal types and terms in blue when they may contain the indicated ``unquote'' forms, discussed later.}
\label{syntax-IL}
%\vspace{-10px}
\end{figure}

Programs in @$\lambda$ are written as \emph{external terms}, $e$. The concrete syntax of external terms is essentially conventional, as suggested by the left side of the example in Figure \ref{example}. It desugars (purely syntactically) to a subtantially more uniform abstract syntax, shown in Figure \ref{syntax-EL} and discussed as we go on. We review recent techniques that permit modularly introducing new desugarings like these in Sec. \ref{desugaring}.

The static and dynamic semantics are specified simultaneously by a \emph{bidirectionally typed translation semantics} targeting an \emph{internal language}, with terms $\iota$. The two key judgements have the form:  \[\esynX{e}{\st\moutput}{\iota\moutput} ~~~~~\text{and}~~~~~ \eanaX{e}{\st}{\iota\moutput}\]
\noindent

These are pronounced ``$e$ (synthesizes / analyzes against) type $\sigma$ with translation $\iota$ under typing context $\Upsilon$ and tycon context $\Phi$''. We indicate ``outputs'' when introducing judgement forms by a \emph{mode annotation}, shown; these are not part of the judgement's syntax. The rules, shown in Figure \ref{statics-EL}, will be discussed incrementally as we explain the example in Figure \ref{example}.  Bidirectional typechecking is also sometimes called \emph{local type inference} \cite{Pierce:2000:LTI:345099.345100}. %We will see how we leverage this bidirectionality when we discuss literal forms below. % The judgement is only well-defined for \emph{valid} tycon contexts, written judgementally as $\vdash \Phi$ and \emph{valid} typing contexts, $\vdash_\Phi \Upsilon$, which we also specify below.
 \begin{figure*}[t!]
\small
\hspace{-5px}\input{example-table}
%\\~\\
%$\begin{array}{rcl}
%e\cdot\conclbl{lbl}(\conclbl{lbl}_1=e_1, \cdots, \conclbl{lbl}_n=e_n)  :=  \etarg{(\conclbl{lbl}; \svar{list}n[\klbl]~\conclbl{lbl}_1~\cdots~\conclbl{lbl}_n)}{e}{e_1; \cdots; e_n}
%\end{array}$
\caption{An example external term, $e_{ex}$, in concrete syntax (left), desugared to abstract syntax (right), with static terms shown in green in examples (only). The notation ${\small \scolor{\desugar{...}}}$ stands for an embedding of the indicated ${\small \scolor{\conclbl{label}}}$, ${\small \scolor{\concrx{regular expression}}}$, ${\small \scolor{\concstr{string}}}$ or numeral into the SL, with derived kinds  $\klbl$, $\krx$,  $\kstr$ and $\knat$, not shown. The signatures of helper functions, e.g. ${\small \scolor{\svar{nil}}}$ and ${\small \scolor{\svar{list}n}}$, are shown in Figure \ref{helper-sigs}.}
\label{example}
\end{figure*}

\paragraph{Internal Language} @$\lambda$ relies on a \emph{typed internal language} with support for type abstraction. In this paper, we use {$\mathcal{L}\{{\rightharpoonup}\,{\forall}\,{\mu}\,{\iunit}\,{\times}\,{+}\}$} (Figure \ref{syntax-IL}), though our results do not depend on this precise choice of internal type constructors, only on the general forms of the judgements. We assume the internal statics are specified in the standard way by judgements for internal type formation {$\iDelta~\iTheta \vdash \tau$}, typing context formation { $\iDelta \vdash \iGamma$} and type assignment {$\iDelta~\iGamma \vdash \iota : \tau\moutput$}. The internal dynamics are also a standard structural operational semantics with judgements {\small $\iota \mapsto \iota\moutput$} and {$\iota~\mathtt{val}$} (cf. \cite{pfpl}).

We also here define a syntax for simultaneous substitutions, of terms for variables, $\gamma$, and of types for polymorphic type variables, $\delta$, and assume standard judgements ensuring that these are valid with respect to a valid context, $\Delta \vdash \gamma : \Gamma$ and $\vdash \delta : \Delta$. We apply these substitutions to terms, types and typing contexts using the syntax $[\gamma]\iota$, $[\delta]\iota$, $[\delta]\tau$ and $[\delta]\Gamma$ (in some prior work, application of a substitution like this is written using a hatted form, e.g. $\hat\gamma(\iota)$; we intend the same semantics but use a notation more consistent with standard substitution, e.g. $[\iota/x]\iota'$). We will omit  leading $\emptyset$ and $\cdot$ in examples; in specifications, the former is used for  metatheoretic finite mappings, the latter for metatheoretic ordered lists.

This style of specification has some  similarities to the Harper-Stone \emph{typed elaboration semantics} for Standard ML \cite{Harper00atype-theoretic}. There, however, external and internal terms were governed by a common type system. In @$\lambda$, {internal types} classify only internal terms. 
This style may thus also be compared to specifications for the first stage of a type-directed compiler, e.g. the TIL compiler for Standard ML \cite{tarditi+:til-OLD}, here lifted ``one level up'' into the language. We will see how the techniques developed to reason about typed compilation are lifted into our semantics to compositionally reason about type safety, which requires that the translation of every well-typed external term is a well-typed internal term, below.

\paragraph{Kinds and Static Terms}
\begin{figure}[t!]
\small
\hspace{-5px}\input{syntax-table-EL}\\
\caption{Abstract syntax of the external language (EL). Terms in gray are technical devices with no corresponding concrete syntax.}
\label{syntax-EL}
\end{figure}
\begin{figure}[t!]
\small
\hspace{-4px}\input{syntax-table-SL}
\caption{Syntax of the static language (SL).}
\label{syntax-SL}
\end{figure}
The workhorse of our calculus is the \emph{static language}, which itself forms a typed lambda calculus where 
%External terms are classified by (external) \emph{types}. 
\emph{kinds}, $\kappa$, classify \emph{static terms}, $\sigma$.  The syntax of the SL is given in Figure \ref{syntax-SL}. We note that the core of the SL is a total functional programming language based on several standard fragments: total functions, quantification over kinds, inductive kinds (constrained by a positivity condition to prevent non-termination), and products and sums (these closely follow \cite{pfpl} so we again omit some details).

The kinding judgement, specified  in Figure \ref{statics-SL}, takes the form $\sofkX{\st}{\kappa\moutput}$, where $\kDelta$ and $\kGamma$ are analagous to $\iDelta$ and $\iGamma$ (and analagous well-formedness judgements are defined, also omitted). The natural number $n$ is used as a technical device in our metatheory to prevent terms of the form $\sana{n}{\st}$ and $\ssyn{n}$ from arising if not permitted; these forms, discussed later, never need to be written directly (they would have no corresponding concrete syntax), so $n$ can assumed 0 for all user code. $\kXi$ is described below. The normalization judgement (specified in Figure \ref{dynamics-SL}) takes the form $\snorm{\st}{\memD}{\memG}{\aCtx}{\st\moutput}{\memD\moutput}{\memG\moutput}$, where $\memD$, $\memG$ and $\aCtx$ are also technical devices that will be described later; they too can be ignored from the perspective of user code. We write $\st \Downarrow \st'$ iff $\snorm{\st}{\emptyset}{\cdot}{{\rightharpoonup}; \emptyset; \emptyset}{\st'}{\emptyset}{\cdot}$. \emph{Static values} are normalized static terms. Normalization can also raise an error (to indicate a type error in an external term, or a problem in a tycon definition, as we will discuss), indicated by the judgement $\serrX{\st}$. We omit error propagation rules.%We will refer to the relevant rules as we proceed.

 %We will specify the statics (i.e. kinding) and dynamics (i.e. normalization) of the static language in Figures \ref{statics-SL} and \ref{dynamics-SL}.%The normalization judgement takes the form $\snorm{\st}{\memD}{\memG}{\aCtx}{\st\moutput}{\memD\moutput}{\memG\moutput}$.  %This core can be considered  a simply-typed dialect of a language like Coq (we explore this further in Sec. \ref{coq}).

\paragraph{External Types are Static Values}

External types (or simply \emph{types}) are {static values} of kind $\kty$.  %We will return to kinds $\kity$ and $\kitm$ below.
The form $\sty{c}{\st}$ introduces a type by ``applying'' a \emph{tycon}, $c$, to a static term, $\st$, called the \emph{type index}. There is one built-in tycon, $\rightharpoonup$, classifying possibly partial external functions, discussed below. All other types are constructed by tycons defined in the tycon context and identified by name. We write tycon names in small caps, e.g. $\tcvar{lprod}$, and use $\tc$ as a metavariable ranging over these (Figure \ref{syntax-TC}). We will return to the other introductory form for types, $\sotherty{m}{\tau}$, which also serves only as a technical device, later. %\todo{mention otherty here?}

Two user-defined types are constructed on line 1 of Figure \ref{example}: a labeled product type declaring a single field labeled $\conclbl{venue}$ of type $\tcvar{rstr}~\concrx{([A-Z]+) \digit\{4\}}$, which classifies \emph{regular strings}, known statically to be in the regular language of the indicated regular expression, which can contain nested parenthesized groups. %We write these indices using conventional concrete syntax. 
%The regular expression can contain parenthesized groups.% We will discuss the operations these provide below. %Although the concrete syntax resembles function application, in our calculus tycons are not themselves static terms.

\paragraph{Index Kinds} The rule \rulename{k-ty} in Figure \ref{statics-SL} requires that the type index must be of the \emph{index kind} of the tycon, given by the \emph{index kind context}, $\kXi$, a simple mapping from tycons to kinds (Figure \ref{syntax-TC}). The initial index kind context, $\kXi_0$, is always included, defining the index kind of ${\rightharpoonup}$ as $\kty \times \kty$. 
The example index kind context in Figure \ref{example-tcsigs}, $\kXi_\text{ex}$, specifies that the index kind of $\tcvar{lprod}$ is $\klist{\kprod{\klbl}{\kty}}$, and the  index kind of $\tcvar{rstr}$ is $\krx$. We assume $\klist{\kappa}$, $\klbl$ and $\krx$ are embedded into the static language in the standard way. The concrete syntax used on line 1 desugars to static terms of these kinds (this sugar could be defined modularly as \emph{kind-specific syntax} by adapting the technique described in Sec. \ref{desugaring} \cite{TSLs}). If a tycon context, $\Phi$, is valid, it  uniquely determines a valid index kind context, as specified  in Figure \ref{statics-TC} by the judgement $\tccok{\Phi}{\kXi\moutput}$, detailed below. In our example, $\tccok{\Phi_\text{ex}}{\kXi_\text{ex}}$.

\paragraph{Type-Level Computation}
The labeled product type on line 1 is bound to the static variable $\svar{V}$  in the remainder of the external term. This  desugars to the more general external construct $\eslet{\st}{\svar{x}}{e}$, on the right. The rules \rulename{syn-slet} and \rulename{ana-slet}, shown in Figure \ref{statics-EL}  (here, the former applies) begin by assigning a kind to $\sigma$ under the index kind context determined by $\Phi$, then substitute its value into the external term $e$ to continue with synthesis or analysis. Static variables, written in bold, are distinct from external and internal variables. The rule \rulename{n-ty} in Figure \ref{dynamics-SL} shows that normalizing $\sty{c}{\st}$ simply involves normalizing $\st$. Because types are static values but need not be written directly in normal form, this is  a form of \emph{type-level computation of higher kind}, though our SL has a more general role than a standard type-level language. %Type indices (e.g. common regular expressions) and, as we will shortly discuss, term indices can also be computed.%, so our use of a static language type-level computation.

\begin{figure}[t]
\small
\hspace{-5px}$\begin{array}{rcllcl}
\multicolumn{3}{l}{\mvlbl{tycons}~~~c~::=~{\rightharpoonup}\,~|~ \tc} & \multicolumn{3}{l}{\mvlbl{index kind contexts}}\\
\multicolumn{3}{l}{\mvlbl{tycon contexts}} & \kXi_0 & := & {\rightharpoonup}[\kty \times \kty]\\
\Phi & ::= & \cdot ~|~ \Phi, \phi & \kXi & ::= & \kXi_0 ~|~ \kXi, \tc[\kappa]\\
\multicolumn{3}{l}{\mvlbl{tycon defs}} & \multicolumn{3}{l}{\mvlbl{tycon sigs}}\\
\phi & ::= & \tcdef{\tc}{\psi}{\opdefs}~~~~& \psi & ::= & \tcsig{\kappa}{\opsigs}\\
\multicolumn{3}{l}{\mvlbl{opcon structures}} & \multicolumn{3}{l}{\mvlbl{opcon sigs}}\\
\opdefs & ::= & \keyw{rep} = \st  & \Omega & ::= & \cdot\\
 & ~|~ & \omega, \analitop{\st} & & ~|~ & \Omega, \litsig{\kappa}\\
 & ~|~ & \opdefs, \synop{\opname{op}}{\st} & & ~|~ & \opsigs, \opsig{\opname{op}}{\kappa}
\end{array}$
\caption{Syntax of tycons and opcons + sigs and contexts.}
\label{syntax-TC}
\end{figure}
%
\begin{figure}
\small
\hspace{-5px}$\begin{array}{l@{\hspace{2px}}c@{\hspace{2px}}l@{\hspace{10px}}l@{\hspace{2px}}c@{\hspace{2px}}l}
\Phi_\text{ex} & := & \phi_\text{rstr}, \phi_\text{lprod} & \kXi_\text{ex} & := &\kXi_0, \tcvar{rstr}[\krx], \tcvar{lprod}[\klist{\klbl\times\kty}]\\
\phi_\text{rstr} & & \text{\normalsize(Figure \ref{tc-rstr})} & \psi_\text{rstr} & := &\tcsig{\krx}{\Omega_\text{rstr}}\\
\phi_\text{lprod}& &\text{\normalsize(Figure \ref{tc-lprod})} & \psi_\text{lprod} &:=& \tcsig{\klist{\klbl \times \kty}}{\Omega_\text{lprod}}\\
\Omega_\text{rstr} & := & \multicolumn{4}{r}{\litsig{\kstr}, \opsig{\opname{concat}}{\kunit}, \opsig{\opname{\#}}{\keyw{Nat}}, \opsig{\opname{coerce}}{\krx}, \opsig{\opname{case}}{\kunit}, \opsig{\opname{!rc}}{\kunit}}\\
\Omega_\text{lprod} & := & \multicolumn{4}{r}{\litsig{\klist{\klbl}}, \opsig{\opname{\#}}{\klbl}, \opsig{\opname{with}}{\klist{\klbl}}, \opsig{\opname{!ri}}{\kunit}}
\end{array}$
\caption{The contexts and signatures for Figures \ref{tc-rstr} and \ref{tc-lprod}.}
\label{example-tcsigs}
\end{figure}

\paragraph{Type Constructor Contexts}

\begin{figure}[t]
\small
~\hfill{\fbox{$\tccok{\Phi}{\kXi\moutput}$}}\vspace{-20px}
\begin{mathpar}
\hspace{-45px}\inferrule[tcc-emp]{ }{\tccok{\cdot}{\kXi_0}}
~~~~~~~~~~~~
\inferrule[tcc-ext]{
  \tccok{\Phi}{\kXi}\\
  \tcok{\kXi}{\phi}{\tc}{\ktyidx}
}{\tccok{\Phi, \phi}{\kXi, \tc[\ktyidx]}}
\end{mathpar}
%
%
~\hfill{\fbox{$\tcok{\kXi}{\phi}{\tc\moutput}{\kappa\moutput}$}}\vspace{-20px}
%
%
\begin{mathpar}
\hspace{-65px}\inferrule[tcdef-ok]{
  \tc \notin \text{dom}(\kXi)~~~~
  \tcsigok{\psi}{\ktyidx}~~~~
  \ocok{\kXi}{\omega}{\tc}{\psi}
}{
  \tcok{\kXi}{\tcdef{\tc}{\psi}{\omega}}{\tc}{\ktyidx}
}
\end{mathpar}
~\hfill{\fbox{$\tcsigok{\psi}{\kappa\moutput}$}}\vspace{-20px}
\begin{mathpar}
\hspace{-50px}\inferrule[tcsig-ok]{
  \keq{\emptyset}{\ktyidx}\\
  \ocsigok{\Omega}
}{
  \tcsigok{\tcsig{\ktyidx}{\Omega}}{\ktyidx}
}
\end{mathpar}
~\hfill\fbox{$\keq{\kTheta\minput}{\kappa\minput}$}\vspace{-16px}\begin{mathpar}
\hspace{-50px}\inferrule[keq-k]{\kvar \in \kTheta}{\keq{\kTheta}{\kvar}}
~~~~~~
\inferrule[keq-ind]{\keq{\kTheta, \kvar}{\kappa}}{
  \keq{\kTheta}{\kmu{\kvar}{\kappa}}}
~~~~~~
\inferrule[keq-unit]{ }{\keq{\kTheta}{\kunit}}
\\
\inferrule[keq-prod]{
  \keq{\kTheta}{\kappa_1}~~~~
  \keq{\kTheta}{\kappa_2}
}{
  \keq{\kTheta}{\kprod{\kappa_1}{\kappa_2}}
}
~~~~~~
\inferrule[keq-sum]{
  \keq{\kTheta}{\kappa_1}~~~~
  \keq{\kTheta}{\kappa_2}
}{
  \keq{\kTheta}{\ksum{\kappa_1}{\kappa_2}}
}
~~~~~~
\inferrule[keq-ty]{ }{\keq{\kTheta}{\kty}}
\end{mathpar}
~\hfill\fbox{$\ocsigok{\Omega}$}\vspace{-18px}
\begin{mathpar}
\hspace{-20px}\inferrule[ocs-rep]{ }{\ocsigok{\cdot}}
~~~~
\inferrule[ocs-lit]{
  \keq{\emptyset}{\kappa}
}{
  \ocsigok{\litsig{\kappa}}
}
~~~~
\inferrule[ocs-targ]{
  \ocsigok{\Omega}~~~~
  \opname{op} \notin \text{dom}(\Omega)~~~~
  \keq{\emptyset}{\kappa}
}{
  \ocsigok{\Omega, \opsig{\opname{op}}{\kappa}}
}
\end{mathpar}
$\karg := (\kunit \rightarrow (\kty \times \kitm)) \times (\kty \rightarrow \kitm)$~\hfill\fbox{$\ocok{\kXi}{\omega}{\tc}{\psi}$}%\vspace{-18px}
\begin{mathpar}
\inferrule[oc-rep]{
  \sofkz{\emptyset}{\emptyset}{\kXi}{\stx{rep}}{\karrow{\ktyidx}{\kity}}
}{
  \ocok{\kXi}{\keyw{rep}=\stx{rep}}{\tc}{\tcsig{\ktyidx}{\cdot}}
}

\inferrule[oc-lit]{
  \ocok{\kXi}{\omega}{\tc}{\tcsig{\ktyidx}{\Omega}}\\
  \sofkz{\emptyset}{\emptyset}{\kXi, \tc[\ktyidx]}{\stx{def}}{{\ktyidx}\rightarrow{\kappa}\rightarrow{\kargs}\rightarrow {\kitm}}
}{
  \ocok{\kXi}{\omega, \analitop{\stx{def}}}{\tc}{\tcsig{\ktyidx}{\Omega, \litsig{\kappa}}}
}

\inferrule[oc-targ]{
  \ocok{\kXi}{\omega}{\tc}{\tcsig{\ktyidx}{\Omega}}\\
  \sofkz{\emptyset}{\emptyset}{\kXi, \tc[\ktyidx]}{\stx{def}}{{\ktyidx}\rightarrow\kitm\rightarrow{\kappa}\rightarrow{\kargs}\rightarrow{(\kty\times \kitm)}}\\
}{
  \ocok{\kXi}{\omega, \synop{\opname{op}}{\stx{def}}}{\tc}{\tcsig{\ktyidx}{\Omega, \opsig{\opname{op}}{\kappa}}}
}
\end{mathpar}
\caption{Type constructor kinding.}
\label{statics-TC}
\end{figure} A tycon context, $\Phi$, is a list of tycon definitions, $\phi$ (Figure \ref{syntax-TC}). Each tycon definition is annotated with a \emph{tycon signature}, $\psi$, which specifies the index kind of the tycon as well as an \emph{opcon signature}, $\Omega$, which relates to the \emph{opcon structure}, $\omega$, discussed below.  The tycons used in our example are defined in Figures \ref{tc-rstr} and \ref{tc-lprod}.  In Figure \ref{statics-TC}, the judgement $\tcok{\kXi}{\phi}{\tc\moutput}{\ktyidx\moutput}$ ensures that their names are unique, their signatures are valid, and that the opcon structure respects the opcon signature.%, ``returning'' the information needed to extend $\kXi$.

% \begin{figure}[t]
% \small
% $\begin{array}{rcl}
% \karg & := & (\kunit \rightarrow (\kty \times \kitm)) \times (\kty \rightarrow \kitm)\\
% %\svar{arity0} : \klist{\karg} \rightarrow \kunit & \svar{syn} : \karg \rightarrow \kty \times \kitm\\
% %\svar{arity1} : \klist{\karg} \rightarrow \karg & \svar{ana} : \karg \rightarrow \kty \rightarrow \kitm\\
% %\svar{arity2} : \klist{\karg} \rightarrow \karg \times \karg
% \end{array}$
% \caption{The external semantics exposes arguments as list of pairs of static functions for synthesis and analysis, respectively. We assume the two helper functions above in the examples.}
% \label{karg}
% \end{figure}
\begin{figure}[t]
\small
%\renewcommand{\slam}[3]{\lambda #2{:}#1.#3}
\hspace{-5px}$\begin{array}{l}
\tcdef{\tcvar{rstr}}{\psi_\text{rstr}}{\\
\quad\keyw{rep}=\scolorc\slam{\krx}{\svar{tyidx}}{\sqity{\qcolorc\imu{s}{\keyw{str}\times\keyw{list}[s]}\scolorc}\color{black}}\\
% ana lit
\quad \analitop{
  \scolorc\slam{\krx}{\svar{tyidx}}{\slam{\kstr}{\svar{tmidx}}{\slam{\kargs}{\svar{args}}{\cnl
\quad\quad \keyw{let}~\svar{aok} : \kunit = \svar{arity0}~\svar{args}~\keyw{in}~\cnl
\quad\quad \keyw{letpair}~(\svar{chars}, \svar{groups})  = \svar{rmatch}~\svar{tyidx}~\svar{tmidx}~\keyw{in}~\cnl
\quad\quad\quad \sqitm{\qcolorc\ifold{s}{\keyw{str}\times\keyw{list}[s]}{\cnl\qcolorc
\quad\quad\quad\quad (\quq{\scolorc\svar{str\_of\_Str}~\svar{chars}\qcolorc},\quq{\scolorc\svar{rstrs\_of\_Strs}~\svar{groups}\qcolorc})}\scolorc}}}}
}\color{black},\\
% syn concat
\quad \synop{\opname{concat}}{\scolorc\slam{\krx}{\svar{ttyidx}}{\lambda \svar{ttr}{:}\kitm.\slam{\kunit}{\svar{tmidx}}{\slam{\kargs}{\svar{args}}{\cnl
\quad\quad \keyw{letpair}~(\svar{targ}, \svar{a})  =\svar{arity2}~\svar{args}~\keyw{in}~\keyw{letpair}~(\svar{aty}, \svar{atr})=\svar{syn}~\svar{a}~\keyw{in}\cnl
\quad\quad \stycase{\tcvar{rstr}}{\svar{aty}}{\svar{atyidx}}{(\sty{\tcvar{rstr}}{\svar{rconcat}~\svar{ttyidx}~\svar{atyidx}},\cnl
\quad\quad\quad \sqitm{\qcolorc rsconcat~\quq{\scolorc\svar{ttr}\qcolorc}~\quq{\scolorc\svar{atr}\qcolorc}\scolorc})}{\sraise{\kty \times \kitm}}}}}}\color{black},\\
% syn case
%\quad \synop{\opname{case}}{\slam{\krx}{\svar{ttyidx}}{\slam{\kitm}{\svar{ttr}}{\slam{\kunit}{\svar{tmidx}}{\slam{\kargs}{\svar{args}}{\\
%\quad\quad \svar{arity2}~\svar{args}~(\slam{\karg}{\svar{a1}}{\slam{\karg}{\svar{a2}}{\svar{syn}~\svar{a1}~\slam{\kty}{\svar{a1ty}}{\slam{\kitm}{\svar{a1tr}}{\\
%\quad\quad\quad \keyw{let}~\svar{hdty} : \kty =\sty{\tcvar{rstr}}{\svar{rxhd}~\svar{ttyidx}}~\keyw{in}\\
%\quad\quad\quad \keyw{let}~\svar{tlty} : \kty =\sty{\tcvar{rstr}}{\svar{rxtl}~\svar{ttyidx}}~\keyw{in}\\
%\quad\quad\quad \keyw{let}~\svar{a2ty} : \kty =\sparr{(\svar{hdty}, \sparr{(\svar{tlty}, \svar{a1ty})})}~\keyw{in}\\
%\quad\quad\quad \svar{ana}~\svar{a2}~\svar{a2ty}~\slam{\kitm}{\svar{a2tr}}{
%        (\svar{a1ty}, \\
%\quad\quad\quad\quad \sqitm{\iAp{\qtuq{\srep{\svar{a1ty}}}}{scase}~\quq{\svar{ttr}}~\ilam{\iunit}{\_}{\quq{\svar{a1tr}}}~\quq{\svar{a2tr}}})}}}}}}}}}}\\
%% syn eq
%\quad \synop{\opname{eq}}{\slam{\krx}{\svar{ttyidx}}{\slam{\kitm}{\svar{ttr}}{\slam{\kunit}{\svar{tmidx}}{\slam{\kargs}{\svar{args}}{\\
%\quad\quad \svar{arity3}~\svar{args}~\slam{\karg}{\svar{a1}}{\slam{\karg}{\svar{a2}}{\slam{\karg}{\svar{a3}}{\\
%\quad\quad\quad \svar{syn}~\svar{a1}~\slam{\kty}{\svar{a1ty}}{\slam{\kitm}{\svar{a1tr}}{\stycase{\tcvar{rx}}{\svar{a1ty}}{\svar{a1tyidx}}{\\
%\quad\quad\quad \svar{synsame}~\svar{a2}~\svar{a3}~\slam{\kty}{\svar{bty}}{\slam{\kitm}{\svar{a2tr}}{\slam{\kitm}{\svar{a3tr}}{(\svar{bty},\\
%\quad\quad\quad\quad \sqitm{\iAp{\qtuq{\srep{\svar{bty}}}}{sequal}~\quq{\svar{ttr}}~\quq{\svar{a1tr}}~\\
%\quad\quad\quad\quad\quad \ilam{\iunit}{\_}{\quq{\svar{a2tr}}}~\ilam{\iunit}{\_}{\svar{a3tr}}})}}}}{err}}}}}}}}}}}\\
% syn grp
\quad \synop{\opname{\#}}{\scolorc\slam{\krx}{\svar{ttyidx}}{\slam{\kitm}{\svar{ttr}}{\slam{\knat}{\svar{tmidx}}{\slam{\kargs}{\svar{args}}{\cnl
\quad\quad \keyw{let}~\svar{aok} : \kunit =\svar{arity0}~\svar{args}~\keyw{in}\cnl
\quad\quad \keyw{let}~\svar{rg} : \krx = \svar{rgroupn}~\svar{ttyidx}~\svar{tmidx}~~\keyw{in}\cnl
\quad\quad\quad (\sty{\tcvar{rstr}}{\svar{rg}}, \sqitm{\qcolorc rsgroupn~\quq{\scolorc\svar{nat\_of\_Nat}~\svar{tmidx}\qcolorc}~\quq{\scolorc\svar{ttr}\qcolorc}\scolorc})
}}}}}\color{black},\\
% syn coerce
\quad \synop{\opname{coerce}}{\scolorc\slam{\krx}{\svar{ttyidx}}{\slam{\kitm}{\svar{ttr}}{\slam{\krx}{\svar{tmidx}}{\slam{\kargs}{\svar{args}}{\cnl
\quad\quad \keyw{let}~\svar{aok} : \kunit = \svar{arity0}~\svar{args}~\keyw{in}\cnl
\quad\quad \keyw{let}~\svar{slok} : \kunit = \svar{rsublang}~\svar{tmidx}~\svar{ttyidx}~\keyw{in}\cnl
\quad\quad \keyw{let}~\svar{rtr} : \kitm = \svar{rx\_of\_Rx}~\svar{tmidx}~\keyw{in}\cnl
\quad\quad \quad (\sty{\tcvar{rstr}}{\svar{tmidx}},\sqitm{\qcolorc rsregroup~\quq{\scolorc\svar{rtr}\qcolorc}~\quq{\scolorc\svar{ttr}\qcolorc}\scolorc})}}}}}\color{black},\\
% syn check
\quad \synop{\opname{case}}{\scolorc\slam{\krx}{\svar{ttyidx}}{\slam{\kitm}{\svar{ttr}}{\slam{\kunit}{\svar{tmidx}}{\slam{\kargs}{\svar{args}}{\cnl
\quad\quad \keyw{letpair}~(\svar{a1}, \svar{a2})=\svar{arity2}~\svar{args}~\keyw{in}~\keyw{letpair}~(\svar{a1ty}, \svar{a1tr})=\svar{syn}~\svar{a1}~\keyw{in}~\cnl
\quad\quad \sparrcase{\svar{a1ty}}{\svar{a1tyidx}}{\keyw{letpair} (\svar{a1tyin}, \svar{a1tyout}) = \svar{a1tyidx}~\keyw{in}\cnl
\quad\quad \stycase{\tcvar{rstr}}{\svar{a1tyin}}{\svar{r}}{\keyw{let}~\svar{rtr}:\kitm = \svar{rx\_of\_Rx}~\svar{r}~\keyw{in}\cnl
\quad\quad \keyw{let}~\svar{a2tr} : \kitm = \svar{ana}~\svar{a2}~\svar{a1tyout}~\keyw{in}\cnl
\quad\quad\quad (\svar{a1tyout}, \sqitm{\qcolorc\iAp{\qtuq{\scolorc\srep{\svar{a1tyout}}\qcolorc}}{rscheck}~\quq{\scolorc\svar{rtr}\qcolorc}~\quq{\scolorc\svar{ttr}\qcolorc}~\cnl\qcolorc
\quad\quad\quad\quad \quq{\scolorc\svar{a1tr}\qcolorc}~\ilam{\iunit}{\_}{\quq{\scolorc\svar{a2tr}\qcolorc}}\scolorc})\cnl
\quad\quad}{\sraise{\kty \times \kitm}}}}{\sraise{\kty \times \kitm}}}}}}\\
\quad \synop{\opname{!rc}}{\scolorc \lambda \svar{ttyidx}{:}\krx.\lambda \svar{ttr}{:}\kitm.\lambda \svar{tmidx}{:}{\kunit}.\lambda \svar{args}{:}\kargs.\cnl
\quad\quad (\sqitm{\sty{\tcvar{rstr}}{\desugar{\concrx{\digit+}}}}, \svar{str\_of\_Str}~{\desugar{\concstr{13}}})}\color{black}}
\end{array}
$
\caption{The definition of the $\tcvar{rstr}$ tycon, $\phi_\text{rstr}$.}
\label{tc-rstr}
\end{figure}
\begin{figure}[t]
\small
\hspace{-5px}$\begin{array}{l}
\tcdef{\tcvar{lprod}}{\psi_\text{lprod}}{\\
\quad\keyw{rep}=\scolorc\lambda\svar{tyidx}{:}\klist{\klbl \times \kty}.\skap{\kity}{\skap{\klbl\times\kty}{\svar{listrec}}}~\svar{tyidx}~\sqity{\qcolorc\iunit\scolorc}~\cnl
  \quad\quad (\lambda \svar{h}{:}\klbl \times \kty.\lambda \svar{r}{:}\kity.\keyw{letpair}~(\_,\svar{hty})=\svar{h}~\keyw{in}~\cnl
  \quad\quad\quad \sqity{\qcolorc\qtuq{\scolorc\svar{r}\qcolorc}\times\qtuq{\scolorc\srep{\svar{hty}}\qcolorc}\scolorc})\color{black},\\
\quad\analitop{\scolorc\lambda\svar{tyidx}{:}\klist{\klbl\times\kty}.\lambda\svar{tmidx}{:}\klist{\klbl}.\lambda\svar{args}{:}\kargs.\cnl
  \quad\quad \keyw{let}~\svar{inhabited}:\kunit=\svar{uniqmap}~\svar{tyidx}~\keyw{in}\cnl
  \quad\quad \skap{\kitm}{\skap{\karg}{\skap{\klbl}{\skap{\klbl\times\kty}{\svar{listrec3}}}}}}~\svar{tyidx}~\svar{tmidx}~\svar{args}~\sqitm{\qcolorc\itriv\scolorc}\cnl
  \quad\quad\quad \lambda \svar{h1}{:}\klbl\times\kty.\lambda \svar{h2}{:}\klbl.\lambda\svar{h3}{:}\karg.\lambda \svar{r}{:}\kitm.\cnl
  \quad\quad\quad\quad \keyw{letpair}~(\svar{rowlbl}, \svar{rowty})=\svar{h1}~\keyw{in}~\keyw{let}~\svar{lok}:\kunit=\svar{lbleq}~\svar{rowlbl} ~\svar{h2}~\keyw{in}\cnl
  \quad\quad\quad\quad \keyw{let}~\svar{rowtm} : \kitm = \svar{ana}~\svar{h3}~\svar{rowty}~\keyw{in}~\sqitm{\qcolorc(\quq{\scolorc\svar{r}\qcolorc}, \quq{\scolorc\svar{rwtm}\qcolorc})\scolorc}\color{black},\\
\quad\synop{\opname{\#}}{\scolorc\lambda \svar{ttyidx}{:}\klist{\klbl\times\kty}.\lambda \svar{ttr}{:}\kitm.\lambda\svar{tmidx}{:}\klbl.\lambda\svar{args}{:}\kargs.\cnl
  \quad\quad \keyw{let}~\svar{aok} : \kunit = \svar{arity0}~\svar{args}~\keyw{in}\cnl
  \quad\quad \keyw{letpair}~(\svar{rownum}, \svar{rowty})=\svar{lookup}~\svar{ttyidx}~\svar{tmidx}~\keyw{in}\cnl
  \quad\quad\quad (\svar{rowty}, \svar{prjnth}~\svar{rownum}~\svar{ttr}))}\color{black},\\
\quad\synop{\opname{with}}{\scolorc\cnl
\quad\lambda \svar{ttyidx}{:}\klist{\klbl\times\kty}.\lambda \svar{ttr}{:}\kitm.\lambda\svar{tmidx}{:}\klist{\klbl}.\lambda\svar{args}{:}\kargs.\cnl
  \quad\quad \keyw{letpair}~(\svar{olist}, \svar{otr})=\skap{\klist{\klbl\times\kty}\times\kitm}{\skap{\karg}{\skap{\klbl}{\svar{listrec2}}}}~\cnl
  \quad\quad\quad \svar{tmidx}~\svar{args}~(\svar{ttyidx}, \svar{ttr})\cnl
  \quad\quad\quad (\lambda \svar{h1}{:}{\klbl}.\lambda\svar{h2}{:}{\karg}.\lambda\svar{r}{:}{\klist{\klbl\times\kty}\times\kitm}.\cnl
  \quad\quad\quad\quad \keyw{letpair}~(\svar{rowty}, \svar{rowtr})=\svar{syn}~\svar{h2}~\keyw{in}~\keyw{letpair}~(\svar{rlist}, \svar{rtr})=\svar{r}~\keyw{in}\cnl
  \quad\quad\quad\quad\quad (\skap{\klbl\times\kty}{\svar{append}}~\svar{rlist}~(\svar{h1}, \svar{rowty}),\cnl
  \quad\quad\quad\quad\quad \sqitm{\qcolorc(\quq{\scolorc\svar{rtr}\qcolorc}, \quq{\scolorc\svar{rowtr}\qcolorc})\scolorc}))~\keyw{in}~\cnl
  \quad\quad \keyw{let}~\svar{inhabited}:\kunit=\svar{uniqmap}~\svar{olist}~\keyw{in}~\cnl
  \quad\quad (\sty{\tcvar{lprod}}{\svar{olist}}, \svar{otr})}\\
\quad\synop{\opname{!ri}}{\scolorc \lambda \svar{ttyidx}{:}\krx.\lambda \svar{ttr}{:}\kitm.\lambda \svar{tmidx}{:}{\kunit}.\lambda \svar{args}{:}\kargs.\cnl
\quad\quad (\sqitm{\sty{\tcvar{rstr}}{\desugar{\concrx{\digit+}}}}, \sqitm{\qcolorc\ifold{s}{\keyw{str}\times\keyw{list}[s]}{\\
\quad\quad\quad \qcolorc(\quq{\scolorc\svar{str\_of\_Str}~\desugar{\concstr{oops}}\qcolorc},\quq{\scolorc\svar{rstrs\_of\_Strs}~\skap{\kstr}{\svar{nil}}\qcolorc})}\scolorc})}\color{black}}
\end{array}$
\caption{The definition of the $\tcvar{lprod}$ tycon, $\phi_\text{lprod}$.}
\label{tc-lprod}
\end{figure}
\paragraph{Type Equivalence} 
To simplify the handling of type equivalence, type index kinds must be \emph{equality kinds}: those for which semantic equivalence coincides with syntactic equality at normal form. We define these by the judgement $\keq{\kDelta}{\kappa}$, appearing as a premise of \rulename{tcsig-ok} in Figure \ref{statics-TC}. Equality kinds are similar to equality types as found in Standard ML. The main implication of this choice is that type indices cannot contain static functions%, an issue we will return to in Sec. \ref{variants} when discussing polymorphic and recursive  types. %Type equality is not an extensible part of the calculus, an issue that we will return to below. %We implicitly identify terms up to alpha renaming of bound variables throughout this paper, as is the usual convention (algorithmically, a locally nameless representation can be used).

\paragraph{Subsumption} The rule \rulename{subsume}, Figure \ref{statics-EL}, simply states that if a term synthesizes a type, it can be analyzed against that type. This relies on the fact that equivalent types are syntactically identical, and a metatheoretic property that states that the type synthesis judgement, assuming well-formed contexts, actually synthesizes a type (stated formally in  Section \ref{calculus}). Subsumption will only be applied once in our example, on line 7, discussed below.

\paragraph{Representations}
\begin{figure}
\small
\small
~\hfill\fbox{$\sfinalrepX{\st}{\tau\moutput}$}\vspace{-18px}
\begin{mathpar}
\inferrule[conc-rep]{
  \boxdot = \text{names}(\Phi)\\\\
  \snorm{\srep{\st}}{\emptyset}{\cdot}{\boxdot,{\rightharpoonup}; \emptyset; \Phi}{\sqity{\ity}}{\memD}{\cdot}\\
  \vdash_\Phi \memD \leadsto \delta : \Delta\\
  \Delta~\emptyset \vdash \tau
}{
  \sfinalrep{\Phi}{\st}{[\delta]\tau}
}
\end{mathpar}
~\hfill\fbox{$\sfinalrepX{\Upsilon}{\Gamma\moutput}$}\vspace{-18px}
\begin{mathpar}
\hspace{-75px}\inferrule[trans-ctx-emp]{ }{
  \sfinalrep{\Phi}{\emptyset}{\emptyset}
}
~~~~~~~~
\inferrule[trans-ctx-ext]{
  \sfinalrep{\Phi}{\Upsilon}{\Gamma}\\
  \sfinalrep{\Phi}{\st}{\tau}
}{
  \sfinalrep{\Phi}{\Upsilon, x \Rightarrow \st}{\Gamma, x : \tau}
}
\end{mathpar}
~\hfill\fbox{$\vdash_\Phi \memD \leadsto \delta\moutput : \Delta\moutput$}\vspace{-18px}
\begin{mathpar}
\hspace{-74px}\inferrule[D-emp]{ }{\emptyset \leadsto \emptyset : \emptyset}
~~~~~
\inferrule[D-ext]{
  \vdash_\Phi \memD \leadsto \delta : \Delta\\
  \vdash_\Phi \st \leadsto \tau
}{
  \vdash_\Phi \memD, \st \leftrightsquigarrow \alpha \leadsto \delta, \tau/\alpha : \Delta, \alpha
}
\end{mathpar}
\caption{Concrete Representations}
\label{representations}
\end{figure}
Each external type has a corresponding internal type called its \emph{concrete representation}. Each tycon definition specifies the concrete representations of the types it constructs with a \emph{representation schema}, specified as the first component of the opcon structure. Rule \rulename{oc-rep} specifies that this is simply a static function mapping type indices to static terms of kind $\kity$, which are introduced by the quotation form $\sqity{\qcolorc\tau\color{black}}$. Unquote forms, written $\qcolorc\qtuq{\color{black}\st\qcolorc}$, normalize away  recursively, shown in rules \rulename{n-qity-*}.

In Figure \ref{tc-rstr}, the representation schema of $\tcvar{rstr}$ simply ignores the type index, mapping all regular string types to the same internal type: a recursive type pairing a string  with a list of other regular strings, corresponding to the ``pre-extracted'' parenthesized groups (internal types $\keyw{str}$ and $\keyw{list}[\tau]$ are defined in the usual way, and are distinguished from analagous kinds $\keyw{Str}$ and $\keyw{List}$ by initial capitalization). This is of course not the only valid choice of representation -- one could simply use $\keyw{str}$ (increasing the cost of group extraction, discussed below), or use a tree of offsets, or inspect the index to extract the number of groups, permitting the use of nested tuples (in practice, the IL might provide fixed length vectors). %This should remind readers of the situation of choosing different representation types for an abstract type declared my a module signature. As we will see, the internal representation of a type constructed in this way is similarly held abstract outside of the opcon structure, so this choice is up to the tycon definition.

In Figure 9, the representation schema of $\tcvar{lprod}$ constructs a simple nested tuple representation based on the type index (the labels are not needed in the translation). Because the type index itself contains types, the representation schema must refer to the representations of these types. The static language provides the operator $\srep{\st}$ to extract knowledge about the representation of type $\st$. The kinding rule \rulename{k-rep} is simple, but the normalization semantics are more interesting. Rule \rulename{n-rep-conc}  extracts the representation schema and  passes it the type index, but it does not always apply. For types constructed by a tycon other than ones named in the \emph{concrete rep whitelist}, $\boxdot$ (a simple list of tycons), an \emph{abstract representation}, is generated. The first time a representation is requested, a fresh internal type variable, $\alpha$, is generated and the correspondence is recorded in the \emph{abstract rep store}, $\memD$, as seen in rule \rulename{n-rep-abs}. Later requests return this type variable. Requesting the representation is a simple effect. Given $\svar{V}$ from our example and abbreviating the index of the type of the field $\conclbl{venue}$ as $\stx{rsidx}$: \[\scriptsize\snorm{\srep{\normalsize\svar{V}\scriptsize}}{\emptyset}{\cdot}{{\rightharpoonup},\tcvar{lprod}; \emptyset; \Phi_\text{ex}}{\sqity{\iunit\times\alpha}}{\emptyset, \sty{\tcvar{rstr}}{\stx{rsidx}} \leftrightsquigarrow \alpha}{\cdot}\]

The judgement $\vdash_\Phi \memD \leadsto \delta\moutput : \Delta\moutput$ constructs a substitution and corresponding internal type variable context by generating the concrete representations for each stored type. For the above example: \[\vdash_{\Phi_\text{ex}} \emptyset, \sty{\tcvar{rstr}}{\stx{rsidx}} \leftrightsquigarrow \alpha \leadsto \emptyset, \imu{s}{\keyw{str} \times \keyw{list}[s]}/\alpha : \emptyset, \alpha\]

%We will see how this underlies our results in Sec. \ref{calculus}.
The judgement $\sfinalrepX{\st}{\tau\moutput}$ extracts the concrete representation by placing every tycon in $\Phi$ in the whitelist and checking that the representation is well-formed (${\rightharpoonup}$ is always in the whitelist; $\memD$ may still be non-empty because of $\sotherty{m}{\tau}$, also discussed later):\[\snorm{\srep{\svar{V}}}{\emptyset}{\cdot}{{\rightharpoonup},\tcvar{rstr},\tcvar{lprod}; \emptyset; \Phi_\text{ex}}{\sqity{\iunit\times\imu{s}{\keyw{str}\times\keyw{list}[s]}}}{\emptyset}{\cdot}\]% The representations of the types in Figure \ref{example} need not be known to understand its semantics.

\paragraph{Variables}

\begin{figure*}[t]
\small
~\hfill\fbox{$\esynX{e}{\st\moutput}{\iota\moutput}$}
~\fbox{$\eanaX{e}{\st}{\iota\moutput}$}
\begin{mathpar}
\inferrule[subsume]{
  \esynX{e}{\st}{\iota}
}{
  \eanaX{e}{\st}{\iota}
}

\inferrule[syn-slet]{
  \tccok{\Phi}{\kXi}\\
  \sofkz{\emptyset}{\emptyset}{\kXi}{\st}{\kappa}\\\\
  \snormT{\st}{\st'}\\
  \esynX{[\st'/\sx]e}{\stx{ty}}{\iota}
}{
  \esynX{\eslet{\st}{\sx}{e}}{\stx{ty}}{\iota}
}

\inferrule[ana-slet]{
  \tccok{\Phi}{\kXi}\\
  \sofkz{\emptyset}{\emptyset}{\kXi}{\st}{\kappa}\\\\
  \snormT{\st}{\st'}\\
  \eanaX{[\st'/\sx]e}{\stx{ty}}{\iota}
}{
  \eanaX{\eslet{\st}{\sx}{e}}{\stx{ty}}{\iota}
}

\inferrule[ascribe]{
  \tccok{\Phi}{\kXi}\\
  \sofkz{\emptyset}{\emptyset}{\kXi}{\st}{\kty}\\\\
  \snormT{\st}{\st'}\\
  \eanaX{e}{\st'}{\iota}
}{
  \esynX{\easc{e}{\st}}{\st'}{\iota}
}

\inferrule[syn-var]{
  x \Rightarrow \st \in \Upsilon
}{
  \esynX{x}{\st}{x}
}

\inferrule[syn-let]{
  \esynX{e_1}{\st_1}{\iota_1}\\\sfinalrepX{\st_1}{\tau_1}\\\\
  \esyn{\Upsilon, x \Rightarrow \st_1}{\Phi}{e_2}{\st_2}{\iota_2}\\
}{
  \esynX{\elet{e_1}{x}{e_2}}{\st_2}{\iap{\ilam{\tau_1}{x}{\iota_2}}{\iota_1}}
}

\inferrule[ana-let]{
  \esynX{e_1}{\st_1}{\iota_1}\\\sfinalrepX{\st_1}{\tau_1}\\\\
  \eana{\Upsilon, x \Rightarrow \st_1}{\Phi}{e_2}{\st_2}{\iota_2}\\
}{\eanaX{\elet{e_1}{x}{e_2}}{\st_2}{\iap{\ilam{\tau_1}{x}{\iota_2}}{\iota_1}}}

\inferrule[ana-fix]{
  \sfinalrepX{\st}{\tau}\\\\
  \eana{\Upsilon, x \Rightarrow \st}{\Phi}{e}{\st}{\iota}
}{
  \eanaX{\efix{x}{e}}{\st}{\ifix{\tau}{x}{\iota}}
}

\inferrule[syn-lam]{
  \tccok{\Phi}{\kXi}\\
  \sofkz{\emptyset}{\emptyset}{\kXi}{\st_1}{\kty}\\
  \snormT{\st_1}{\st_1'}\\
  \sfinalrepX{\st_1'}{\tau_1}\\\\
  \esyn{\Upsilon, x \Rightarrow \st_1'}{\Phi}{e}{\st_2}{\iota}
}{
  \esynX{\elam{\st_1}{x}{e}}{\sparr{(\st_1',\st_2)}}{\ilam{\tau_1}{x}{\iota}}
}

\inferrule[syn-ap]{
  \esynX{e_1}{\sparr{(\st_1, \st_2)}}{\iota_1}\\\\
  \eanaX{e_2}{\st_2}{\iota_2}
}{
  \esynX{\eap{e_1}{e_2}}{\st_2}{\iap{\iota_1}{\iota_2}}
}

\inferrule[ana-lit]{
  \tcdef{\tc}{\tcsig{\ktyidx}{\Omega}}{\omega} \in \Phi\\\\
  \litsig{\klitidx} \in \Omega\\
  \tccok{\Phi}{\kXi}\\
  \sofkz{\emptyset}{\emptyset}{\kXi}{\stmidx}{\klitidx}\\\\
  \analitop{\stx{def}} \in \omega\\
  \mkargsX\\\\
  \snorm{\stx{def}~\sttyidx~\stmidx~\stx{args}}{\emptyset}{\memG_0}{{\rightharpoonup}, \tc; \Upsilon; \Phi}{\sqitm{\itm_\text{abs}}}{\memD}{\memG}\\\\
  \snorm{\srep{\sty{\tc}{\sttyidx}}}{\memD}{\cdot}{{\rightharpoonup}, \tc;\emptyset;\Phi}{\sqity{\ity_\text{abs}}}{\memD'}{\cdot}\\\\
  \vdash_\Phi \memD' \leadsto {\delta} : \iDelta\\
  \memG \leadsto {\gamma} : \iGamma\\
  \iDelta~\iGamma \vdash \iota_\text{abs} : \tau_\text{abs}
}{
  \eanaX{\elit{\stmidx}{\es}}{\sty{\tc}{\sttyidx}}{[{\delta}][{\gamma}]\iota_\text{abs}}
}

\inferrule[syn-targ]{
  \esynX{e_\text{targ}}{\sty{\tc}{\sttyidx}}{\iota_\text{targ}}\\\\
  \tcdef{\tc}{\tcsig{\ktyidx}{\Omega}}{\omega} \in \Phi\\\\
  \opsig{\opname{op}}{\ktargidx} \in \Omega\\
  \tccok{\Phi}{\kXi}\\
  \sofkz{\emptyset}{\emptyset}{\kXi}{\stmidx}{\ktargidx}\\\\
  \synop{\opname{op}}{\stx{def}} \in \omega\\
  \mkargs{e_\text{targ}; \es}{\stx{args}}{\memG_0}{n}\\\\
  %\snorm{\srep{\sty{\tc}{\sttyidx}}}{\emptyset}{\cdot}{{\rightharpoonup}, \tc;\emptyset;\Phi}{\sqity{\ity_\text{targ}}}{\memD_0}{\cdot}\\\\
  \snorm{\stx{def}~\sttyidx~(\svar{tr}~\ssyn{0})~\stmidx~(\svar{atl}~\stx{args})}{\emptyset}{\memG_0}{{\rightharpoonup},\tc; \Upsilon; \Phi}{(\st, \sqitm{\itm_\text{abs}})}{\memD}{\memG}\\\\
  \snorm{\srep{\st}}{\memD}{\cdot}{{\rightharpoonup},\tc;\emptyset;\Phi}{\sqity{\ity_\text{abs}}}{\memD'}{\cdot}\\\\
  \vdash_\Phi \memD' \leadsto {\delta} : \iDelta\\
  \memG \leadsto {\gamma} : \iGamma\\
  \iDelta~\iGamma \vdash \iota_\text{abs} : \tau_\text{abs}
}{
  \esynX{\etarg{\opname{op}}{\sttmidx}{e_\text{targ}}{\es}}{\st}{[\delta][\gamma]\iota_\text{abs}}
}

\inferrule[ana-other]{
  \sfinalrep{\Phi}{\srep{\Upsilon}}{\Gamma \parallel \memD}\\
  \memD \leadsto \delta; \Delta\\
  \Delta~\Gamma \vdash \iota : \tau
}{
  \eanaX{\eother{\iota}}{\sotherty{n}{\tau}}{[\delta]\iota}
}
\end{mathpar}
\caption{Typing}
\label{statics-EL}
\end{figure*}

We can now continue to line 2 of Figure \ref{example}, where we bind a term to a variable using the form $\elet{e_1}{x}{e_2}$. The rules \rulename{syn-let} and \rulename{ana-let} give an essentially standard semantics. The translation is to lambda application. The concrete representation of the synthesized type for $e_1$ determines the type annotation. The \rulename{syn-var} rule is standard. The well-formedness judgement for external typing contexts, $\ctxok{\Phi}{\Upsilon}$, ensures that variables map to types (Figure \ref{statics-EL2}). The judgement $\vdash_\Phi \Upsilon \leadsto \Gamma\moutput$ translates the types in $\Upsilon$ to their concrete representations (Figure \ref{representations}).

%The rule \rulename{ana-fix} similarly permits fixpoint definitions in analytic positions (this fixpoint operator follows Plotkin's PCF, cf. \cite{pfpl}). 
\paragraph{Functions}

On lines 4-6, we see a function definition. The syntax and semantics of functions are fixed and built into the external language (so that the semantics can uniformly handle the tricky issue of variable binding). Only the argument types need to be provided explicitly; the return type is synthesized, as shown in the rules \rulename{syn-lam}. Rule \rulename{syn-ap} is standard. Function types are written concretely as $\sigma_1 \rightharpoonup \sigma_2$ and abstractly as $\sparr{(\sigma_1, \sigma_2)}$, i.e. the index kind of the built-in type constructor $\rightharpoonup$ is $\kprod{\kty}{\kty}$ (cf. $\kXi_0$ in Figure \ref{example-tcsigs}). By the semantics of the operators described below, the full type synthesized by $paper$, written concretely, will be:
\[\begin{array}{l}
\concty{rstr}{\concrx{..*}} \rightharpoonup \concty{rstr}{\concrx{\digit\digit*}} \rightharpoonup \concty{labeledprod}{\{\\
\quad \conclbl{venue} : \concty{rstr}{\concrx{([A-Z]+) \digit+}}, \conclbl{title} : \concty{rstr}{\concrx{..*}},\\
\quad \conclbl{doi} : \concty{rstr}{\concrx{\digit\digit\rdot\digit\digit\digit\digit\textbackslash/\digit\digit*}}\}}
\end{array}\]
 Like internal functions (and unlike static functions), external functions may be partial, via a $\keyw{fix}$point operator. Rule \rulename{ana-fix} follows Plotkin's PCF, cf. \cite{pfpl}, here defined analytically. %(an analytic lambda could also be defined).

\paragraph{Ascriptions} Type ascriptions allow the programmer to explicitly specify the type a term should be analyzed against. For analytic terms appearing in synthetic positions, an ascription is necessary. In the abstract syntax, type ascriptions are always placed directly on terms using the form $\easc{e}{\sigma}$ (in the concrete syntax, they may appear on the variable being bound for convenience). The rule \rulename{syn-asc} in Figure \ref{statics-EL} specifies that the ascription must be a closed static term of kind $\kty$ under the index kind context determined by the tycon context. It is normalized to a type before proceeding with analysis. This is also an essentially standard rule.

\paragraph{Literals} The variable $v$ binds a term introduced \emph{literally}, under an ascription specifying that it should be analyzed against type $\svar{V}$. Inside this literal, we also see a literal, though with no ascription (it will be analyzed against the type specified in $\svar{V}$, as discussed below). All literal forms in the concrete syntax desugar to the same abstract form, $\elit{\stmidx}{\es}$. The \emph{term index},  $\stmidx$, captures statically known portions of the literal as a single static term and the \emph{argument list}, $\es$,  captures all external sub-terms. For example, in the labeled product literal, desugaring constructs a list of field labels as the term index (using standard helper functions whose signatures are shown in Figure \ref{helper-sigs}) and passes in the  corresponding field values as arguments. For regular string literals, the string is lifted into the SL as the term index. There are no sub-terms, so the argument list is empty.

Literal terms can only appear in analytic positions in our calculus. More specifically, they are given meaning by {the type constructor of the type they are being analyzed against}. 
This tycon's opcon signature, $\Omega$,  determines the \emph{literal term index kind} governing $\stmidx$. In Figure \ref{example-tcsigs}, $\psi_\text{lprod}$ specifies $\Omega_\text{lprod}$ which in turn specifies the literal term index kind $\klist{\klbl}$, and $\psi_\text{rstr}$ similarly specifies $\Omega_\text{rstr}$ and ${\kstr}$. The literal index kind must also be an equality kind, as specified by rule \rulename{ocs-lit} in Figure \ref{statics-TC}. The rule governing type analysis of literals is \rulename{ana-lit}, shown in Figure \ref{statics-EL}. It begins by checking the provided term index against this kind (premises 1-4).% Note that here, the index kind context tracks only type index kinds, as operators are not themselves constructed as static values \todo{mention future work thing?}.

\subparagraph{Literal Opcon Definitions}

The next premise of \rulename{ana-lit} extracts the \emph{literal opcon definition}, $\stx{def}$, defined using the form $\analitop{\stx{def}}$ in the opcon structure, $\omega$, of the tycon definition. This  is responsible for deciding the translation of the literal as a function of the type index, the term index and  the types and translations of the arguments. %The judgement $\ocok{\kXi}{\omega}{\tc}{\psi}$, in Figure \ref{statics-TC}, checks the opcon definition list $\omega$ for tycon $\tc$ against its signature $\psi$.

The relevant rule in Figure \ref{statics-TC} is \rulename{oc-lit}, which ensurses that $\stx{def}$ has kind $\ktyidx \rightarrow \klitidx \rightarrow \kargs \rightarrow \kitm$. The kinds $\ktyidx$ and $\klitidx$ are determined by the signature as just described. The kind $\kitm$ is analagous to $\kity$, and has one introductory ``quotation'' form, $\sqitm{\qcolor{\iota}}$. Normalization produces an internal term without such forms (rules \rulename{n-qitm-*} in Figure \ref{dynamics-SL}), i.e. a standard IL term. Note that these can only be opaquely composed, not inspected (there is no elim form) and $\kity$ and $\kitm$ are not equality kinds.

Before discussing arguments and the remaining premises, let us turn to the literal opcon definition for $\tcvar{rstr}$ in Figure \ref{tc-rstr}, which will be called by the rule \rulename{ana-lit} to derive a translation for the field value on line 1, $e_\text{venue} := \elit{\desugar{\concstr{EXMPL 2015}}}{\cdot}$: \footnotesize\[\begin{array}{l}\eana{\emptyset}{\Phi_\text{ex}}{e_\text{venue}}{\sty{\tcvar{rstr}}{\desugar{\concrx{([A-Z]+) \digit+}}}}{\ifold{s}{\keyw{str}\times\keyw{list}[s]}{\\
\quad \keyw{pair}(\llfloor\concstr{EXMPL 2015}\rrfloor; \iAp{\imu{s}{\keyw{str}\times\keyw{list}[s]}}{list1}~\ifold{s}{\keyw{str}\times\keyw{list}[s]}{\\
\quad\quad \keyw{pair}(\llfloor\concstr{EXMPL}\rrfloor; \iAp{\imu{s}{\keyw{str}\times\keyw{list}[s]}}{nil})})}}
\end{array}\]
%where we define the representation \footnotesize$\tau_{rs} := \imu{s}{\keyw{s} \times \keyw{list}[s]}$
\normalsize
We write $\llfloor{...}\rrfloor$ for the embedding of the indicated string into the IL, and abbreviate the definitions of $\keyw{str}$, $\keyw{list}[s]$ and the helper functions for working with lists as above. We refer to the translation as $\iota_\text{venue}$.

The logic in Figure \ref{tc-rstr} simply checks that the argument list is empty using the simple helper function $\svar{arity0}$, which raises an error if the list is non-empty (rules \rulename{k-raise} and \rulename{n-raise}; in practice, the provider would here provide an explicit error message to the client). It then calls the helper function $\svar{rmatch}$ on the term index to make sure it matches the regular expression provided as the type index, also extracting its groups statically. Finally, it constructs the translation above by calling simple helper functions for lowering static strings and lists into the IL using quotations. We do not show the details; this is, by design, exactly the code one would write in a compiler written in a simply-typed functional language for the branch of the case analysis in the translation logic corresponding to the regular string literal term constructor. Here, we have pulled it into the tycon definition, pushed the case analysis into the semantics, and do not require a separate term constructor by relying on bidirectional typechecking to permit us to share syntax amongst many types.% (see Sec. \ref{desugaring}).%Reasoning about the correctness of this function requires specifying stronger invariants than the kind system provides, and our purpose in this section is not to claim to guarantee extension correctness. Our point is that mistakes will be caught

\paragraph{Arguments} Let us now return to the important issue of how the semantics handles arguments by discussing how the literal opcon definition for $\tcvar{lprod}$ in Figure \ref{tc-lprod} is called by \rulename{ana-lit} to derive the following translation (assuming the definition of $\svar{list1}$):
\footnotesize\[\begin{array}{l}
\stx{tmidx} := \skap{\klbl}{\svar{list1}}~\desugar{\conclbl{venue}}\\
\stx{tyidx} := \conckap{\svar{list1}}{\kprod{\klbl}{\kty}}~(\desugar{\conclbl{venue}}, \sty{\tcvar{rstr}}{\stx{rsidx}})\\
\eana{\emptyset}{\Phi_\text{ex}}{\elit{\stx{tmidx}}{e_\text{venue}}}{\sty{\tcvar{lprod}}{\stx{tyidx}}}{\ipair{\itriv}{\iota_\text{venue}}}
\end{array}\]
\normalsize
Note that we did not here exploit the opportunity to optimize the representation of labeled products of length 1. Our interest here is not in this particular decision,  but to ensure that such decisions can be freely explored and have only local implications, like changes in the representation of an abstract type in an ML-like module system.

To call the opcon definition, the semantics needs to construct the \emph{argument interface}, $\stx{args}$, a list of type $\klist{\karg}$. %It also constructs an \emph{argument interface list}, $\stx{args}$ as well as an \emph{argument store}, $\memG_0$, from the argument list, $\es$, as specified by the judgement $\mkargsX$ in Figure \ref{statics-EL2}. The judgement also reports the lengths of these, $n$.
The kind $\karg$ is defined in Figure \ref{statics-TC}: it classifies a pair of static functions that serve as hooks into the semantics, permitting the opcon definition to request synthesis or analysis, respectively, for the argument these functions abstract.  The judgement $\mkargs{\es}{\st\moutput}{\memG\moutput}{n\moutput}$, specified in Figure \ref{statics-EL2}, constructs a list of such function pairs by wrapping the static operators $\ssyn{n}$ and $\sana{n}{\st}$. The kinding rules \rulename{k-syn} and \rulename{k-ana} (and the signatures in Figure \ref{helper-sigs}) imply that $\sofkn{\emptyset}{\emptyset}{\kXi}{n}{\stx{args}}{\kargs}$, where $n$ is the number of arguments. All user-defined static terms, as we have seen, are always checked with $n=0$, ensuring that these operators cannot arise directly in well-kinded opcon definitions,  avoiding the possibility of ``out of bounds'' problems. Passing the argument interface into opcon definitions, which were checked with $n=0$, does not cause kind safety problems due to a simple lemma, provable inductively:
\begin{lemma}
If $\sofkn{\kDelta}{\kGamma}{\kXi}{n}{\st}{\kappa}$ and $n' > n$ then $\sofkn{\kDelta}{\kGamma}{\kXi}{n'}{\st}{\kappa}$.
\end{lemma}

The arguments are placed in an \emph{argument store}, $\memG$. Initially, $\memG_0$ is simply a numbered sequence of the arguments themselves, here $\memG_0 := \cdot, 0 \hookrightarrow e_\text{venue}$. The normalization semantics for $\ssyn{n}$ and $\sana{n}{\st}$ can have an effect on the argument store.

The literal opcon constructor for $\tcvar{lprod}$ in Figure \ref{tc-lprod} begins by ensuring that the type the literal is being analyzed against might actually be inhabited by ensuring that there are no duplicate labels in the type index.\footnote{Implementing a $\keyw{Map}[\kappa]$ kind that ensures this by construction, while remaining an equality kind, requires some simple additions to the SL, e.g. abstract equality kinds modeled also on SML, which we omit for concision and to demonstrate this more generally applicable technique.} It then recurses simultaneously over the term index, type index and argument list, checking corresponding labels for equality and requesting analysis of each argument against the corresponding type (the helper function $\svar{ana}$ is in Figure \ref{helper-sigs}; $\svar{listrec3}$ raises an error if the inputs are of different length). The rule \rulename{n-ana} shows how the argument is extracted from the store and analyzed in the typing context and tycon context provided (by rule \rulename{ana-lit} here) in the \emph{operation context}, $\aCtx$. If it succeeds, the argument store is updated with the type provided for analysis, the translation and, crucially, a fresh internal variable, $x$, standing for its translation. The abstract representation of the type provided for analysis is also produced using the whitelist in $\aCtx$, which contains only the tycon of the type the literal is being analyzed against (and the function tycon, which is never held abstract so functions can be used for operations with binding structure, e.g. \opname{case}, below). We write $\memG \otimes n \hookrightarrow ...$ to indicate that the previous value in the store for $n$ is removed. Because $\tcvar{rstr}$ is not in $\boxdot$, we will thus have:\footnotesize\[\begin{array}{l}
\snorm{\stx{rep}~\sttmidx~\sttyidx~\stx{args}}{\emptyset}{\memG_0}{{\rightharpoonup},\tcvar{lprod}; \emptyset; \Phi_\text{ex}}{\sqitm{\ipair{\itriv}{x}}}{\memD}{\memG}\\
\memD := \emptyset, \sty{\tcvar{rstr}}{\stx{rsidx}} \leftrightsquigarrow \alpha\\
\memG := \cdot, 0 \hookrightarrow e_\text{venue} : \sty{\tcvar{rstr}}{\stx{rsidx}} \leadsto \iota_\text{venue}/x : \alpha
\end{array}
\]\normalsize

\paragraph{Abstract Translation Checking} We must now check that this \emph{abstract translation} is \emph{representationally consistent}. The next two premises in \rulename{ana-lit} generate the abstract representation of the type provided for analysis, as well as its corresponding substitution, $\delta$, and internal type variable context, $\Delta$, as was shown previously. The judgement $\memG \leadsto \gamma\moutput : \Gamma\moutput$ generates a term substitution and internal typing context based on $\memG$ similarly (Figure \ref{statics-EL2}). Here, $\memG \leadsto (\emptyset, \iota_\text{venue}/x) : (\emptyset, x : \alpha)$ is generated. Finally, the semantics checks the \emph{abstract translation} above against the abstract representation using these contexts, e.g. here $\emptyset, \alpha~\emptyset, x : \alpha \vdash \ipair{\itriv}{x} : \iunit \times \alpha$. It is easy to see that this holds. Only after performing this check are the substitutions $\delta$ and $\gamma$ applied. We will  show an example where this check fails even when the check would have succeeded if we had applied the substitutions first below  and return to the powerful metatheoretic implications of this check in Sec. \ref{calculus}.

\paragraph{Hygiene} Note that the variables in $\Upsilon$ are not made available in $\Gamma$ when performing this check, guaranteeing \emph{hygienic translation}: the translation must not have any free variables other than those generated implicitly via the argument store (which applying the  substitution will eliminate) so inadvertent capture cannot occur. We assume also that substitution application is \emph{capture-avoiding} (i.e. the binding structure of terms is maintained, as can be implemented straightforwardly using a locally nameless representation). Thus, terms can be reasoned about without conditions on the context that translation logic may place them under, i.e. \emph{compositionally}.

\paragraph{Targeted Operations}

%To better motivate the normalization semantics of $\ssyn{n}$ and $\sana{n}{\st}$ in Figure \ref{dynamics-SL}, which update the argument store, let us look more closely at our examples.

\begin{figure}
~\hfill\fbox{$\ctxok{\Phi}{\Upsilon}$}\vspace{-18px}
\begin{mathpar}
\hspace{-35px}\inferrule[ectx-emp]{ }{
  \ctxok{\Phi}{\cdot}
}
~~~~~~
\inferrule[ectx-ext]{
  \vdash \Phi : \kXi\\
  \sofkz{\emptyset}{\emptyset}{\kXi}{\st}{\kty}\\
  \snormT{\st}{\st}
}{
  \ctxok{\Phi}{\Upsilon, x \Rightarrow \st}
}
\end{mathpar}
~\hfill\fbox{$\mkargs{\es}{\st\moutput}{\memG\moutput}{n\moutput}$}\vspace{-20px}
\begin{mathpar}
\hspace{-20px}\inferrule[mkargs-z]{
  \snormT{\skap{\karg}{\svar{nil}}}{\stx{args}}
}{
  \mkargs{\cdot}{\stx{args}}{\cdot}{0}
}

\inferrule[mkargs-s]{
  \mkargs{\es}{\stx{args}}{\memG}{n}\\
  \snormT{\skap{\karg}{\svar{push}}~\stx{args}~((\lambda \svar{\_}{:}\kunit.\ssyn{n}), (\lambda \svar{t}{:}\kty.\sana{n}{\svar{t}}))}{\stx{args}'}
}{
  \mkargs{\es, e}{\stx{args}}{\memG, n \hookrightarrow e}{n+1}
}
\end{mathpar}
~\hfill\fbox{$\memG \leadsto \gamma\moutput : \Gamma\moutput$}\vspace{-20px}
\begin{mathpar}
\hspace{-40px}\inferrule[G-emp]{ }{\cdot \leadsto \emptyset : \emptyset}
~~~~~~~~
\inferrule[G-ignored]{\memG \leadsto \Gamma}{
  \memG, n \hookrightarrow e \leadsto \Gamma
}

\inferrule[G-ext]{
  \memG \leadsto \gamma : \Gamma
}{
  (\memG, n \hookrightarrow e : \st \leadsto \iota/x : \tau) \leadsto (\gamma, \iota/x) : (\Gamma, x : \tau)
}
\end{mathpar}
\caption{Auxiliary judgements for external statics.}
\label{statics-EL2}
\end{figure}

On line 5 of Figure \ref{example}, we invoke the \emph{labeled product extension opcon}, $\opname{with}$, on $v$ to create a new labeled product with additional fields, $\conclbl{title}$ and $\conclbl{doi}$, with types consistent with those shown in the return type of $paper$ above. The $\conclbl{doi}$ field's value is computed by invoking the \emph{regular string concatenation opcon}, $\opname{concat}$. A number of other similar \emph{targeted operations} are seen elsewhere. These all desugar to the same abstract form, $\etarg{\opname{op}}{\stx{tmidx}}{e_\text{targ}}{\es}$, where metavariable $\opname{op}$ ranges over opcon names, $\stx{tmidx}$ again captures all statically known portions of the operation (e.g. the field names on line 5), $e_\text{targ}$ is the \emph{target} expression, and $\es$ contains all other external subterms.

In our calculus, all targeted operations are synthetic. The type constructor of the type recursively synthesized for the target is delegated responsibility over the semantics of the targeted operation. It must define an \emph{opcon term index kind} in its opcon signature and an \emph{opcon definition} in its opcon structure matching the name provided, and satisfying the kinding conditions in rules \rulename{ocs-targ} and \rulename{oc-targ} of Figure \ref{statics-TC}. The rule \rulename{syn-targ} in Figure \ref{statics-EL} shows that type synthesis and translation of targeted operations shares much in common with the logic for literals just described. The term index is first checked against the term index kind, then the opcon definition is extracted and an argument interface is constructed as before. We include the target itself as the notional first argument, though for convenience we don't require that each opcon re-synthesize the translation by passing the components in as shown. 
Whereas literal opcon definitions only had to decide a translation, targeted opcon definitions must decide both a type and a translation, returning a pair of kind $\kty \times \kitm$. The abstract translation is checked against the abstract representation of the type produced, as  above.% Note that it might have a tycon other than the one delegated control.

We will not describe each opcon in Figures \ref{tc-rstr} and \ref{tc-lprod} in detail. Once again, the code is essentially identical to the code one would write in a standard typechecker. The ``magic'' happens in the dynamics of the SL, only manifesting when there is a problem. We note the following interesting features in these examples:
\begin{itemize}
\item In \opname{concat}, the definition uses the elimination form for types, $\stycase{c}{\st}{\sx}{\st}{\st}$, to extract the type index, compute the concatenated regular expression and construct a translation that performs the concatenation. This implies that tycons in our calculus have features of both open sum types (a default case is required) and modules. Tycon indices are not held abstract. In the calculus as given, there is rarely a reason to inspect the index of a different tycon. If we added support for direct calls between opcons (e.g. so that the regular expression could expose an operator that returned a labeled product containing the groups directly), having this ability would become more useful.
\item The \opname{coerce} operation supports converting between regular string types known statically to obey a sublanguage relation (cf. \cite{conf/cpp/DoczkalKS13}). Rearrangement of group boundaries can also be performed in this way (this requires a regrouping operation in this choice of translation). We invoke it on line 7 in an analytic position, so subsumption is applied. An interesting direction for future work is to add support for describing implicit coercions between types of the same tycon by enriching the subsumption rule.
\item The \opname{case} operation supports checked conversions. The first argument must synthesize a function type, where the index of the input type determines the regular expression that the string will dynamically be checked against. The ``else branch'', which is necessary here but not for a coercion, is analyzed against the return type of this function. Note that on line 9, the else branch will never be taken, but the type system ``forgot'' the information that could be used to optimize it away.
\item Both example tycons define a \opname{\#} opcon, though with different index kinds. They are seen being used on line 8 to extract a field and then a subgroup. Like literals, types are used to distinguish the semantics despite the shared syntax. This has similarities to operator overloading, but here the static and dynamic semantics themselves are being overloaded statically. %Subgroup extraction together with regrouping via coercion and case analysis obviates the need for more conventional elim forms.
\item It is enlightening to derive the abstract translation and representation produced when typechecking $p\opname{\#}\conclbl{venue}$ on line 8, recalling that $p$ is a labeled product with three fields each of a different regular string type, with \conclbl{venue} being the first. Assuming \keyw{rs} is the concrete representation of regular strings: $\begin{array}{l}
\iota_\text{abs} := \keyw{letpair}(x; \_, y.\keyw{letpair}(y; z, \_.z))\\
\tau_\text{abs} := \alpha_1\\
\delta := \emptyset, \keyw{rs}/\alpha_1, \keyw{rs}/\alpha_2, \keyw{rs}/\alpha_3\\
\Delta := \emptyset, \alpha_1, \alpha_2, \alpha_3\\
\gamma := \emptyset, p/x\\
\Gamma := \emptyset, x : \iunit \times (\alpha_1 \times (\alpha_2 \times \alpha_3))
\end{array}$
\item The $\opname{!rc}$ opcon definition shows an example where the representational consistency check would fail simply because a term of a type other than the one determined by the representation schema of $\tcvar{rstr}$ was generated. The $\opname{!ri}$ opcon definition is more interesting, however: the translation produced matches the concrete type of $\tcvar{rstr}$, so type safety would not be violated if it were permitted and it would pass the check described above if it were performed post-substitution. But it is quite problematic for reasons of modular reasoning: it claims that the translation has a regular string type, but the string itself does not obey the \emph{value invariant} that $\tcvar{rstr}$ locally maintains: that the string in the translation is actually in the language specified by the type index. Luckily, it does not pass the abstract translation check because it is not \emph{representationally independent} ($\keyw{fold}$ cannot have type $\alpha$).
\end{itemize}

% \begin{figure}
% ~\hfill\fbox{$\sfinalrepX{\srep{\Upsilon}}{\Gamma\moutput \parallel \memD\moutput}$}
% \begin{mathpar}
% \inferrule[rep-ctx-emp]{ }{
%   \sfinalrep{\Phi}{\srep{\cdot}}{\cdot \parallel \cdot}
% }

% \inferrule[rep-ctx-ext]{
%   \sfinalrepX{\srep{\Upsilon}}{\Gamma \parallel \memD}\\
%   \snorm{\srep{\st}}{\memD}{\cdot}{{\rightharpoonup}; \emptyset; \Phi}{\sqity{\qity}}{\memD'}{\cdot}\\
%   \qity \uparrow \tau
% }{
%   \sfinalrep{\Phi}{\srep{\Upsilon, x\Rightarrow\st}}{\Gamma, x : \tau \parallel \memD'}
% }
% \end{mathpar}
% \caption{Do I need this?}
% \end{figure}
\section{Metatheory of @$\lambda$}\label{calculus}
Let us now make these intuitions more rigorous. At the outset, a disclaimer: the {ideas} described in the previous section, unifying several disparate threads of research into an evidently powerful core calculus, are intended to serve as the main intellectual contributions of the paper. We can only present the theorems and lemmas below semi-formally. Proving the theorems below completely rigorously given the normalization semantics we have presented is difficult, because it often requires us to reason about  intermediate terms, rather than sub-terms. Rigorous proofs would be most straightforward if we had presented an equivalent structural operational semantics, but this would obscure some important aspects of our presentation. We hope to submit mechanized proofs using a more suitable specification style as an artifact, and have started, but not completed, this work using Coq as of submission.

 \paragraph{Type Safety} The first major issue that we must address is kind safety and type safety. Because the semantics of the EL relies on the kinding and normalization semantics of the SL, and the normalization semantics of the SL relies on the semantics of the EL, this is somewhat tricky.
Let us define $\stype{\st}{\Phi}$ iff $\vdash \Phi \sim \kXi$ and $\svalofk{\st}{\kty}$ and $\snormT{\st}{\st}$. Then, we need: 
\begin{theorem}[Representational Consistency]
If $\tccokX$ and $\vdash_\Phi \Upsilon$ and $\vdash_\Phi \Upsilon \leadsto \Gamma$ then $\emptyset \vdash \Gamma$ and \begin{enumerate}
\item If $\esynX{e}{\st}{\iota}$ then $\stypeX{\st}$ and $\vdash_\Phi \st \leadsto \tau$ and $\emptyset~\Gamma \vdash \iota : \tau$.
\item If $\stypeX{\st}$ and $\eanaX{e}{\st}{\iota}$ then $\vdash_\Phi \st \leadsto \tau$ and    $\emptyset~\Gamma \vdash \iota : \tau$.
\end{enumerate}
\end{theorem}
We must simultaneously prove several important lemmas to show Theorem 1. We assume suitable definitions of judgements $\vdash_\Phi \memD$, which makes sure that $\memD$ stores types, and $\memD~\Upsilon \vdash_\Phi^n \memG$, which ensure that the $n$ facts in $\memG$ are truly valid.
% We can begin by observing that the size of the argument store never changes:

% \begin{lemma}[Argument Store]
% If $\tccokX$ and $\sofk{\emptyset}{\emptyset}{\kXi}{\st}{\kappa}$ and $\snormX{\st}{\st'}$ then $|\memG|=|\memG'|$.
% \end{lemma}

% We can next consider only normalization under an empty argument store, allowing us to exclude the $\keyw{ana}$ and $\keyw{syn}$ forms, which are not used in representation schemas, and prove the following mutually defined lemmas.

Many of the cases require ensuring that the normalization semantics is well-behaved (we discuss canonical forms below):

\begin{lemma}[Static Preservation and Normalization]
If $\tccokX$ and $\sofk{\emptyset}{\emptyset}{\kXi}{\st}{\kappa}$  and $\vdash_\Phi \memD$ and $\memD~\Upsilon \vdash_\Phi^n \memG$  and $\snorm{\st}{\memD}{\memG}{\boxdot; \Upsilon; \Phi}{\st'}{\memD'}{\memG'}$then $\sofkz{\emptyset}{\emptyset}{\kXi}{\st'}{\kappa}$ and $\vdash_\Phi \memD'$ and $\memD'~\Upsilon \vdash_\Phi^n \memG'$ and $\snormT{\st'}{\st'}$.
\end{lemma}
We also need these lemmas about concrete representations.
\begin{lemma}[Concrete Representations]
If $\tccokX$ and $\stypeX{\st}$ and $\vdash_\Phi \st \leadsto \tau$ then $\emptyset~\emptyset \vdash \tau$.
\end{lemma}
\begin{lemma}[Ext. Typing Contexts]
If $\tccokX$ and $\vdash_\Phi \Upsilon$ then \begin{enumerate}
\item If $x \Rightarrow \st \in \Upsilon$ then $\stypeX{\st}$.
\item If $\vdash_\Phi \Upsilon \leadsto \Gamma$ then $\emptyset \vdash \Gamma$.
\end{enumerate}
\end{lemma}
Both these and the cases for literals and targeted forms rely on standard properties of substitutions:

\begin{lemma}[Substitutions] If $\delta : \Delta$  and $\Delta \vdash \gamma : \Gamma$ and $\Delta~\Gamma \vdash \iota : \tau$, then $\emptyset \vdash [\delta]\Gamma$ and $\emptyset~\emptyset \vdash [\delta][\gamma]\iota : [\delta]\tau$, and if $\Delta~\emptyset \vdash \tau$ then $\emptyset~\emptyset \vdash [\delta]\tau$.
\end{lemma} 
We also need the assurance that normalization semantics are updating the store in a way that produces  valid substitutions.

\begin{lemma}[Stores] If $\vdash \Phi \sim \kXi$ and $\vdash_\Phi \Upsilon$ and $\vdash_\Phi \memD$ and $\memD~\Upsilon \vdash_\Phi \memG$ and $\snorm{\st}{\memD}{\memG}{\boxdot; \Upsilon; \Phi}{\st'}{\memD'}{\memG'}$ then $\vdash_\Phi \memD' \leadsto \delta : \Delta$ and $\delta : \Delta$ and $\memG' \leadsto \gamma : \Gamma$ and $\Delta \vdash \gamma : \Gamma$.
\end{lemma}

Ensuring that induction is well-founded is the most complex part of these proofs. We observe that the representation schema for a tycon $\tc$ is kinded under a $\kXi$ that does not include $\tc$ (Figure \ref{statics-TC}). Thus, it cannot request the representation of a type constructed by $\tc$ unless that type is extracted from the index.
The argument store used in typechecking a term contains only subterms and the normalization semantics only attempt to typecheck terms in the argument store. For static terms kinded with $n=0$, the argument store is left unchanged. In other circumstances, its domain is fixed.

%\begin{lemma}[User Terms] If $\sofkz{\emptyset}{\emptyset}{\kXi}{\st}{\kappa}$ and $\snormX{\st}{\st'}$ then $\memG'=\memG$. \end{lemma}
The representational consistency theorem is, in essence, a version of the \emph{type-preserving compilation theorem} developed for the TIL compiler for SML \cite{tarditi+:til-OLD}, here lifted into the language and using user-defined mappings from external types to internal types (governed by the lemmas proved above). It implies type safety (that well-typed programs cannot ``go wrong'') by simply setting $\Upsilon=\emptyset$ and invoking the internal type safety theorem:

\begin{lemma}[Internal Type Safety]
If $\emptyset~\emptyset \vdash \iota : \tau$ then either $\iota~\mathtt{val}$ or $\iota \mapsto \iota'$ such that $\emptyset~\emptyset \vdash \iota' : \tau$.
\end{lemma}

\paragraph{Unicity} It is straightforward to prove that the type and translation of a term is unique if it exists. Each form that does not simply defer to a sub-term is either analytic, or synthetic, never both. This is an important property, as it justifies our ``caching'' behavior for $\memG$ and ensures that typing is deterministic.

\begin{theorem}[Unicity]
If $\tccokX$ and $\vdash_\Phi \Upsilon$  and $\eanaX{e}{\st}{\iota}$ and $\eanaX{e}{\st'}{\iota'}$ then  $\st=\st'$ and $\iota=\iota'$.
\end{theorem}

Note that extending the calculus with analytic targeted operations (e.g. case analysis that does not require synthetic branches) needs care to ensure that unicity is maintained, due to subsumption (i.e. analysis can proceed either via an explicit analysis rule or via a synthesis rule). If we define judgements specifying that subsumption should only proceed if an explicit analysis rule fails, this can be addressed, at the expense of some awkwardness in the specification.

\paragraph{Decidability} A useful property is \emph{decidability of typechecking}. Here, we have suggested a decision procedure using the mode annotations, and by ensuring that normalization cannot diverge. We discussed ensuring that requesting a representation, or requesting analysis and synthesis, will not diverge above. The equality kind condition on type indices subsumes the positivity condition that would be needed to ensure that the elim form for types does not introduce non-termination. We must, however, leave rigorously proving strong normalization of the calculus as future work.

A subtle issue arises in rule \rulename{n-syn-fail} and \rulename{n-ana-fail}. In brackets, we assume a judgement defining the situation where typechecking fails, though we do not inductively define it. In practice, this would not be difficult to determine, though in the specification, it is awkward to write down these cases. If the bracketed premises are omitted, normalization becomes weak (i.e. non-deterministic), but only in that it may fail when it was possible for it to succeed. Similar issues arose in the Harper-Stone semantics for SML \cite{Harper00atype-theoretic}.

\paragraph{Stability}
We can now turn to the situation where the tycon context is extended. Doing so does not affect kinding or normalization.
\begin{theorem}[Stable Kinding and Normalization]
If $\vdash \Phi \sim \kXi$ and $\sofkX{\st}{\kappa}$ and $\vdash \Phi, \phi \sim \kXi, \tc[\kappa']$ then \begin{enumerate}
\item $\sofk{\kDelta}{\kGamma}{\kXi, \tc[\kappa']}{\st}{\kappa}$.
\item If $\snorm{\st}{\memD}{\memG}{\boxdot; \Upsilon; \Phi}{\st'}{\memD'}{\memG'}$ then $\snorm{\st}{\memD}{\memG}{\boxdot; \Upsilon; \Phi, \phi}{\st'}{\memD'}{\memG'}$.
\end{enumerate}
\end{theorem}

Several other similar lemmas for judgements indexed by $\Phi$ are omitted. The consequence of this is that extending $\Phi$ does not affect previously well-typed terms. Note that type systems structured as ``bags of rules'', where extensions can pattern match arbitrarily, would not have this property simultaneously with unicity (i.e. there would be ambiguities). Our method of delegating to a tycon, which remains stable upon extension, is the key to maintaining stability.
\begin{theorem}[Stable Typing]
If $\vdash \Phi \sim \kXi$ and $\vdash_\Phi \Upsilon$ and $\vdash_\Phi \Upsilon \leadsto \Gamma$  and $\vdash \Phi, \phi \sim \kXi, \tc[\kappa']$ then
\begin{enumerate}
\item If $\esynX{e}{\st}{\iota}$ then $\esyn{\Upsilon}{\Phi, \phi}{e}{\st}{\iota}$.
\item If $\eanaX{e}{\st}{\iota}$ then $\eana{\Upsilon}{\Phi, \phi}{e}{\st}{\iota}$.
\end{enumerate}
\end{theorem}

\paragraph{Conservativity}
We now turn to the issue of conservativity, i.e. ensuring that the new terms that become well-typed upon extension of the tycon context cannot violate type invariants maintained in the previous ``closed world''. We saw at the end of Sec. 2 an example of a hazardous situation, where the $\tcvar{lprod}$ tycon attempted to introduce a term of type $\sty{\tcvar{rstr}}{\desugar{\concrx{\digit+}}}$ that did not maintain the invariant that the string in the translation was in the language of the regular expression specified in the type index. Luckily, the representation of this type was held abstract from the perspective of the opcon structure of $\tcvar{lprod}$, so the translation, which would have otherwise been representationally consistent, did not pass the abstract representation check, because it was not \emph{representationally independent}, as enforced by our use of an internal type variable to stand for the representation.

\begin{theorem}[Conservativity]If $\vdash \Phi \sim \kXi$ and $\stypeX{\sty{\tc}{\sttyidx}}$ and we have a value invariant ($\forall e$, if $\eana{\emptyset}{\Phi}{e}{\sty{\tc}{\sttyidx}}{\iota}$ then $\iota \mapsto^{*} \iota'$ and $\iota'~\mathtt{val}$ implies $P(\iota')$) and $\vdash \Phi, \phi \sim \kXi, \tc'[\kappa']$ then ($\forall e$, if $\eana{\emptyset}{\Phi, \phi}{e}{\sty{\tc}{\sttyidx}}{\iota}$ then $\iota \mapsto^{*} \iota'$ and $\iota'~\mathtt{val}$ implies $P(\iota')$).\end{theorem}

Note that this is a stronger statement than stability, as $e$ may not have been well-typed before the extension (e.g. the operation $\opname{!ri}$). Our approach will be to introduce a technical device into the language that allows us to show, however, that there is always some other term with the same type and translation in the original closed world, precisely because  type constructors can only treat one another abstractly. Adding this technical device only weakens the set of properties that can be established benignly, in that one can no longer use exhaustiveness arguments to prove that the default branch of the $\keyw{tycase}$ construct is never taken inside an opcon definition.

\begin{lemma}[Other]
If $\vdash \Phi \sim \kXi$ and $\stypeX{\sty{\tc}{\sttyidx}}$ and $\eana{\Upsilon}{\Phi, \phi}{e}{\sty{\tc}{\sttyidx}}{\iota}$ then there exists an $e_\text{other}$ such that $\eana{\Upsilon}{\Phi}{e}{\sty{\tc}{\sttyidx}}{\iota}$.
\end{lemma}
\begin{proof}We proceed by induction on the typing derivation in the extended context. Most cases proceed by induction, so we omit the details. The interesting cases are for literals and targeted operations.

For $e=\elit{\stmidx}{\es}$, we construct $e_\text{other}=\elit{\stmidx'}{\es'}$:
\begin{itemize}
\item We construct $\stmidx'$ from the value of $\stmidx$ by replacing every sub-term of the form $\sty{\tc'}{\stmidx''}$, where $\tc'$ is the new tycon defined by $\phi$, with $\sotherty{m}{\tau}$, where $\vdash_{\Phi, \phi} \sty{\tc'}{\stmidx''} \leadsto \tau$ and choosing an $m$ uniquely determined by $\stmidx''$ (needed below). The kinding rules then imply that $\sofkz{\emptyset}{\emptyset}{\kXi}{\stmidx'}{\kappa_\text{tmidx}}$, where $\kappa_\text{tmidx}$ is the literal index kind of $\tc$, because no other kinding rules for values depend on $\kXi$. This is why opcon indices were restricted to equality kinds.
\item We construct $\es'$ based on the final argument store, $\memG$. In particular, for every entry in $\memG$ of the form $n \hookrightarrow e : \sty{\tc'}{\stmidx''} \leadsto \iota/x : \alpha$, we append an argument of the form  $\easc{\eother{\iota}}{\sotherty{m}{\tau}}$ to $\es'$, where $m$ is determined by the same mapping as above and $\tau$ is determined via the substitution induced by $\memD$.

The kinding and normalization rules for $\sotherty{m}{\tau}$ imply that its representation is always held abstract, and the $\keyw{tycase}$ form always takes the ``else'' branch. This is precisely the same behavior that $\sty{\tc'}{\sttyidx}$ has, because $\tc' \neq \tc$ and the opcon definition was not kinded knowing of $\tc'$. This implies that, when starting with $\mkargs{\es'}{\stx{args}'}{\memG_0'}{n}$, the derivation of $e_\text{other}=\elit{\stmidx'}{\es'}$ will follow the same path, producing the same abstract translation, and the final argument store will differ only at the entries we modified: the $n$th entry will necessarily be of the form $n \hookrightarrow {\eother{\iota}} : {\sotherty{m}{\tau}} \leadsto \iota/x : \alpha$ where $\alpha$ maps to $\sotherty{m}{\tau}$ in the abstract rep store. As a result, the final substitution $\gamma$ maps every $x$ to its corresponding $\iota$, as in the derivation using extended context (in essence, we made the same translation arise from a different type for arguments that had a type constructed by the new tycon).
\end{itemize}
The argument for targeted operations follows an analagous route. Though a bit circuitous in detail, the essence of the proof is simple: because opcon definitions can only extract indices from  tycons that were already in $\Phi$, and because the representations and translations are held abstract by the semantics, we can simply introduce a generic family of ``other types'' and corresponding constants, each specifying its translation directly. These are distinct from the known types, so they cannot violate their invariants, but permit reasoning generically about future extensions.
\end{proof}
%If ** e <= ty[tc](styidx) ~> i then exists e_other s.t. e <= ty[tc](styidx)

%onstruct a proof that $\eana{\emptyset}{\Phi}{e}{\sty{\tc}{\sttyidx}}{\iota}$ for every proof that $\eana{\emptyset}{\Phi, \phi}{e}{\sty{\tc}{\sttyidx}}{\iota}$.

%These generally rely on monolithic specifications, so that exhaustive induction can be used in proofs. Combining specifications is a manual process. Chlipala and others have suggested that if these proofs are constructed by powerful automated procedures, then the burden of re-proving the metatheory may be decreased \cite{chlipala}. While we do not disagree, this approach comes with no guarantee that the automation procedure will succeed, only anecdotal evidence that the procedures are robust. We aim to render this ``hack'' unnecessary by making it possible to combine  proof terms derived separately directly and automatically. We also aim to provide such guarantees to  implementations, not just their specifications.

\section{Related Work and Discussion}\label{prior-work}

% Note also that type constructors are not themselves static values. To apply them in a curried manner or pass them around in the static language, we can define a static function, e.g. $\slam{\kty}{\svar{t1}}{\slam{\kty}{\svar{t2}}{\sparr{(\svar{t1}, \svar{t2})}}}$ has kind $\karrow{\kty}{\karrow{\kty}{\kty}}$.

%To situate our work, and because we build on some recent results, we begin by summarizing prior approaches that can be considered when a simple isomorphic embedding of a fragment is not evident.

\paragraph{Macros}
If an isomorphic embedding can be established but it is too complex, language-integrated static term rewriting  (``macro'') systems, like Template Haskell \cite{SheardPeytonJones:Haskell-02} and Scala's static macros \cite{ScalaMacros2013}, can help by making it possible to generate ``boilerplate code'' automatically, moving the complexity of computing the embedding of a term into a metaprogram, but the static semantics are fixed. If each metaprogram is invoked like a function, inner macros are expanded before outer macros, and the mechanism enforces abstraction barriers by ensuring that the rewriting logic is hygienic, cannot modify surrounding code, and does not depend on shared state, then compositional reasoning is possible: the meaning of a term depends only on its subterms, not on the specific position it appears in a program.% Complex macros can, however, still be difficult to reason about, so many static macro systems impose further constraints (e.g. Scala recommends using only its ``black box'' macros, which  enforce a function-like typing discipline). These do not permit extensions to the underlying type system.
%Rewriting systems that permit global pattern-based dispatch, however, do not admit strong modular reasoning principles (the same term might match multiple patterns, defined separately, creating ambiguities).

\paragraph{Desugaring}\label{desugaring}
Sometimes, dialects address issues of complexity by introducing new syntax. Indeed, most dialects do build in syntax for a few privileged abstractions (e.g. list literals are nearly ubiquitous, monad comprehensions support a key feature of Haskell). %For example, even though lists can be defined using datatypes, most languages build in special ``literal forms'' that make introducing and pattern matching over lists less tedious. Different dialects may similarly choose to  provide concise forms for working with other types of data (e.g. option types, regular expressions or SQL queries). Types not in the language's standard library, however, must typically use a more uniform syntax. 
%Evidence suggests that this leads programmers to choose less semantically meaningful representations (e.g. representing SQL queries as strings), which can cause a number of issues \cite{TSLs}. 
To make the situation less asymmetric, systems that introduce new  ``desugarings'' atop a base language have been developed. For example, %Camlp4 is a language-external system used by the OCaml community. 
dialects of languages built using Sugar* \cite{erdweg2013framework} allow syntax extensions to be packaged separately and combined using language-integrated declarations. If the desugaring logic obeys the same constraints described above for macros, semantic reasoning can be performed compositionally (and recent work has shown how reasoning about type correctness can be automated \cite{conf/icfp/LorenzenE13}).

These systems do not guarantee that the composition of unambiguous grammars will remain unambiguous (and realistic examples of ambiguous combinations are not uncommon, e.g. XML and HTML). %Every combination of desugarings is thus a  dialect with respect to syntactic reasoning. 
Recent work has made progress in addressing this problem by restricting how syntax extensions can interface with the host syntax. Schwerdfeger and Van Wyk require a globally unique start token and describe checkable conditions pertaining to the follow sets of host language non-terminals to guarantee that extensions can be composed unambiguously \cite{conf/pldi/SchwerdfegerW09}. This suffices for modularly adding new keyword-prefixed forms, but literal forms are awkward to define in this way. Omar et al. describe a language-integrated technique specifically for this purpose, using a technique with parallels to the one we are building on \cite{TSLs}. Literal parsing logic is directly associated with user-defined types, forming \emph{type-specific languages} (TSLs), and local type inference controls invocation of TSL parsing logic, guaranteeing that composition is unambiguous. The mechanism guarantees hygiene and inner terms cannot be inspected, so  fully modular reasoning is possible.

%We assumed these desugaring techniques were available to solve issues strictly related to complexity and convenience.

\paragraph{Optimization and Extensible Compilers}
When an embedding that preserves the static semantics and is sufficiently simple exists, but a different embedding would better preserve a desired cost semantics, term rewriting techniques can also be used to perform specialized ``optimizations'', thus achieving an isomorphic embedding. Care must be taken, however, to ensure that the optimized value is not manipulated directly if the rewriting does not preserve the static semantics (i.e. the rewritten term has a different, less constrained type). Type abstraction can be used directly to achieve this property. \emph{Lightweight modular staging}, for example, uses Scala's support for abstract type members in traits to hide the optimized representations from the program \cite{Rompf:2012:LMS}. As long as optimizations are modularly known to be meaning-preserving (to be sure, a non-trivial proof obligation that benefits from mechanization, as in recent work on extensible compiler optimization \cite{conf/pldi/TatlockL10}), they can be composed arbitrarily. %As long as transformations are meaning-preserving (which can also be shown mechanically), they can be applied in any order. %The metaprograms can modularly be shown not to violate the type system of the host language, Modular reasoning is possible as long as dispatch is explicit (e.g. macros) or based on an unambiguous language mechanism (e.g. trait composition in LMS) and made simpler by the fact that transformations must be type-preserving.

\paragraph{Type Refinement}
%When T, the solution is less clear. 
The techniques above require that an embedding that preserves the static semantics of a desired fragment already exists. When new static distinctions need to be made for values of an existing type, but new primitive operations are not needed, one solution is to develop a system of \emph{type refinements} based on the fragment of interest \cite{Freeman91}. For example, one might refine the type of integers to distinguish negative integers. Most proposals for ``pluggable type systems'' describe such type refinement systems \cite{Brac04a}, which require only additional annotations supplied as comments or metadata (e.g. JavaCOP \cite{Andreae:2006:FIP:1167473.1167479}). %Because the dynamic semantics are fixed,  composition is safe. Although our regular string types could perhaps be approximated with refinements, this does not permit the advanced operations, optimizations and representation independence results we enable.

\paragraph{Language Frameworks}
The most general situation is when exposing a fragment requires defining new types as well as new operators, where the static and dynamic semantics governing these new operators make non-trivial use of  statically valued information. 
We saw labeled product types added, which require a new type constructor and new operator constructors. The regular string type's group projection operator had a specialized cost semantics, so a simple refinement would not be able to approximate it. Scala's type system does not track the necessary invariants, so LMS-like optimizations would also not be appropriate.%For example, projection operators,  written concretely in some dialects of ML as \verb|#label|, can be seen as applications of \verb|#|, the projection operator constructor, to \verb|label|, a static index. The type of an operation like \verb|#label e| (sometimes written postfix as \verb|e#label|) is a function of the index of the type of \verb|e|. The dynamics also needs (some trace of) this information to determine how the operation is evaluated.

%Many of the other examples mentioned above as motivations for new dialects have a similar flavor. We will detail another example in the next section of a type constructor that tracks strings known statically to be in a regular language, using it to dispense with unnecessary run-time checks in cases where they are not strictly necessary (affecting the cost component of the dynamic semantics).

%An important observation is that while it is possible to \emph{implement} records by translation to tuples, or nested binary pairs, or lists, or many other types arising from a language like $\mathcal{L}\{{\rightharpoonup}\,{\forall}\,{\mu}\,{\keyw{1}}\,{\times}\,{+}\}$, it is not possible to isomorphically embed them as such (and thus rely on purely syntactic sugar) because there is no unique inverse mapping from, e.g., unlabeled product types to the corresponding record type (many combinations of labeled products may be implemented using the same unlabeled product). This is a common pattern even in more complex situations: a simple internal language (IL) generally suffices as a target for implementing the dynamics of an external language (EL) with a richer static semantics, but an embedding is not possible. This observation underlies the design of most compilers.

 % like SML, Specifying languages in a manner analagous to how they are usually implemented can simplify reasoning about the core of the language.

% and typed translation semantics (the internal terms have a different type system).% Translation semantics

%These types do not admit an isomorphic embedding using only unlabeled tuples because the field projection family of operators, indexed by labels (field projection is written concretely as \verb|#label(e)|, \verb|e#label| or \verb|e.label| in different languages), must be capable of operating on all values of such types. % with the labels removed would not admit a unique inverse. 
%(each equivalence class of record types could be implemented as an abstract type using tuples at considerable inconvenience but there would still be no way to define a projection operator that operated uniformly over all records having a particular field).

% (i.e. fragments defining families of operations that cannot simply be written as a finite collection of functions, examples of which we will discuss). If there are a finite number of operations, and they can be typechecked without inspecting the structure of the types of the arguments, then an abstract type can be used. new primitive operations (including new variants of existing operations) that require static knowledge of these invariants, however, this too is infeasible. take advantage of the invariants being maintained, and these operations must be usable throughout a program, . %At best, annotations might be written as ``comments'' or metadata for use by these tools.

A variety of \emph{language frameworks} have been developed to make it easier to define new languages, in some cases together with a compiler and other tooling for these languages. Various terms have been used in the literature, including \emph{compiler generators} and \emph{language workbenches}; see \cite{erdweg2013state}. %Some \emph{extensible compilers} are also actually language frameworks, because they permit the introduction of new constructs, not just new meaning-preserving optimizations. The maintainers of many compilers, e.g. Haskell's GHC, treat their compiler as a ``laboratory'' for new fragments, toggled by flags or special comments. \emph{Logical frameworks} focus more specifically on helping language designers mechanize the metatheory of language and logic specifications. T
hese tools can decrease the effort needed to define a new language, but they either do not support forming languages from separately defined fragments or provide few modular reasoning principles that make it possible to reason separately about these fragments. Every combination of fragments is a new dialect, and must be reasoned about monolithically. %Efforts to reason modularly areeed, several problems can come up when fragments are combined, so . 
Let us briefly review difficulties that arise.

%Concrete syntax known to be separately unambiguous might not be unambiguous when combined, as was already discussed above. 
If the abstract syntax needs to be extended to support a new fragment, problems also arise. In monolithic settings, terms can be implemented using finite recursive sums (i.e. term constructors are often implemented as ML-style datatype constructors), but this does not permit extension, so open sum types or products of functions (i.e. objects \cite{conf/oopsla/Aldrich13}) must instead be used. This can present issues when one wishes to modularly define new functionality that should exhaustively cover all terms in the language (e.g. pretty-printers for expressions). Reynolds first identified this problem \cite{Reynolds75} and Wadler named it the \emph{expression problem}. 
%A number of language frameworks (e.g. JastAdd \cite{Ekman:2007:JEJ:1297027.1297029} and Silver \cite{VanWyk:2010:SEA}) use extensible \emph{attribute grammars}. These can be seen as a form of open sum, where attributes correspond to functions performing traversals (more specifically, \emph{catamorphisms} \cite{catamorphisms}\todo{citation / remove?}). %When a new term constructor is added, the logic determining how some existing attributes (e.g. typechecking and translation) should handle the new case is also provided. \emph{Forwarding} can be used to attempt to delegate responsibility over attributes other than those explicitly defined to another term constructor, eventually leading to one in the fixed internal language \cite{VanWyk:2010:SEA}. This can address some aspects of the expression problem, but creates a bigger problem: the internal implementation details  of a fragment are necessarily exposed, violating an abstraction barrier that, as we will discuss, is critical.
In this work, we sidestep the problems of syntax, instead leaving it fixed and relying on a bidirectional type system delegating to a relevant tycon. %For example, projection operators do not require adding a corresponding term constructor to the abstract syntax. Instead, projection is categorized as a \emph{targeted operation}, so  the concrete term \verb|#label e| desugars to an abstract term like $\keyw{targ}[(\desugar{\conclbl{\#}}, \desugar{\conclbl{\texttt{label}}})](e)$ (where $\desugar{\cdot}$ denotes an encoding of constant labels, here a label corresponding to the operator constructor and the field label itself, into the static language). The type constructor of the type synthesized by the target argument, here $e$, determines the semantics of the term. If a new record-like fragment is added (e.g. they give the example of one with prototypic inheritance), it can reuse the same concrete and abstract syntax directly, precluding conflicts and avoiding the need to define the behavior of tools like pretty printers for every new fragment. As in their work on type-specific languages, described above, the semantics of literal forms are controlled by the type constructor of the type the literal is being analyzed against. 
%We will discuss this in the next section.

Once syntactic issues have been addressed, however, there are a host of semantic guarantees  that must be established before a language can be relied upon, as we saw. % The most basic guarantee is \emph{type safety}: that the dynamics are well-defined for all terms accepted as well-typed by the statics, and preserve the statics during evaluation. Each fragment can then build on type safety to establish additional \emph{type invariants} stating properties about all types that it constructs, which  clients can then use to reason about programs (e.g. value induction for eager sum types relies on finiteness of the canonical forms). In practice, typechecking is expected be \emph{deterministic} and \emph{decidable} (i.e. terminating) and, except in circumstances where non-determinism is explicitly exposed to programmers, the dynamic semantics are also expected to be deterministic. %Precise formulations of these properties depend on how the semantics are specified.%, so we will make this more precise below.
Modern language frameworks guarantee few or none of these properties about the dialects they produce. More alarmingly, even when these properties have been established for two dialects  (either in the metatheory or mechanically using a logical framework), and syntactic conflicts are addressed, there is no guarantee that merging the dialects together will conserve these properties. We saw examples of such failures of modularity above. If our goal is to integrate fragment composition into the language, these tools are thus of limited utility. Clients would have to take on the burden of reasoning about these basic properties for every combination of ``libraries'' they chose to import.% Improving this state of affairs, so that it more resembles reasoning about separately defined modules (we assume an ML-style module system), is the general topic of this paper.

\begin{figure*}[t]
\small
\newcommand{\twocol}[1]{\multicolumn{2}{l}{#1}}
\newcommand{\threecol}[1]{\multicolumn{3}{l}{#1}}
\hspace{-5px}$\begin{array}{llll}
\svar{arity0} : \klist{\karg} \rightarrow \kunit & \svar{str\_of\_Str} : \kstr \rightarrow \kitm & \svar{uniqmap} : \klist{\klbl \times \kty}\rightarrow \kunit &  \svar{lbleq} : \klbl \rightarrow \klbl \rightarrow \kunit\\
\svar{arity1} : \klist{\karg} \rightarrow \karg & \svar{rstrs\_of\_Strs} : \klist{\kstr} \rightarrow \kitm & {\svar{lookup} : \klist{\klbl\times\kty}\rightarrow \klbl \rightarrow (\knat \times \kty)}\\
\svar{arity2} : \klist{\karg} \rightarrow \karg \times \karg & \svar{rx\_of\_Rx} : \krx \rightarrow \kitm & \svar{rmatch} : \krx \rightarrow \kstr \rightarrow (\kstr \times \klist{\kstr})\\
\svar{nil} : \kforall{\kalpha}{\klist{\kalpha}} & \svar{nat\_of\_Nat} : \knat \rightarrow \kitm & \svar{rconcat} : \krx \rightarrow \krx \rightarrow \krx & \hspace{-15px}rsconcat : \iparr{\iparr{\keyw{rs}}{\keyw{rs}}}{\keyw{rs}}\\
\svar{list1} : \kforall{\kalpha}{\kalpha \rightarrow \klist{\kalpha}} & \svar{prjnth} : \knat \rightarrow \kitm \rightarrow \kitm & \svar{rgroupn} : \knat \rightarrow \krx \rightarrow \krx &\hspace{-15px} rsgroupn : \iparr{\keyw{nat}}{\iparr{\keyw{rs}}{\keyw{rs}}}\\
\svar{list2} : \kforall{\kalpha}{\kalpha \rightarrow \kalpha \rightarrow \klist{\kalpha}} & \svar{append} : \kforall{\kalpha}{\klist{\kalpha}\rightarrow \kalpha \rightarrow \klist{\kalpha}} & \svar{rsublang} : \krx \rightarrow \krx \rightarrow \kunit & \hspace{-15px}rsregroup : \iparr{\keyw{rx}}{\iparr{\keyw{rs}}{\keyw{rs}}}\\
\twocol{\svar{listrec} : \kforall{\kalpha_1}{\kforall{\kalpha}{\klist{\kalpha_1}\rightarrow \kalpha \rightarrow (\kalpha_1\rightarrow\kalpha\rightarrow\kalpha)\rightarrow\kalpha}}} & rscheck : \iforall{t}{\iparr{\keyw{rx}}{\iparr{\keyw{rs}}{\iparr{(\iparr{\keyw{rs}}{t})}{t}}}}\\
\threecol{\svar{listrec2} : \kforall{\kalpha_1}{\kforall{\kalpha_2}{\kforall{\kalpha}{\klist{\kalpha_1}\rightarrow\klist{\kalpha_2}\rightarrow \kalpha \rightarrow (\kalpha_1\rightarrow\kalpha_2\rightarrow\kalpha\rightarrow\kalpha)\rightarrow\kalpha}}}}\\
\threecol{\svar{listrec3} : \kforall{\kalpha_1}{\kforall{\kalpha_2}{\kforall{\kalpha_3}{\kforall{\kalpha}{\klist{\kalpha_1}\rightarrow\klist{\kalpha_2}\rightarrow \klist{\kalpha_3}\rightarrow\kalpha \rightarrow (\kalpha_1\rightarrow\kalpha_2\rightarrow\kalpha_3\rightarrow\kalpha\rightarrow\kalpha)\rightarrow\kalpha}}}}}\\
\threecol{\svar{syn} := \scolorc\lambda \svar{a}{:}\karg.\keyw{letpair}~(\svar{synfn}, \_) = \svar{a}~\keyw{in}~\svar{synfn}~\striv}\\
\threecol{\svar{ana} := \scolorc\lambda \svar{a}{:}\karg.\keyw{letpair}~(\_, \svar{anafn})=\svar{a}~\keyw{in}~\svar{anafn}}
\end{array}$
\caption{Definitions and signatures of helpers used in the examples.}
\label{helper-sigs}
\end{figure*}

Though we have not mechanically proven these properties for our design, we have given strong evidence that these properties are maintained modularly. That is, we need only establish the semantics of the opcon structures in isolation. %We can then rely on, e.g., the Conservativity theorem above. 
The most immediate direction for future work is to finishing mechanizing the metatheory for this  technique, and to embed it into an existing proof assistant (using a continuation passing style to simulate the stores in our normalization semantics). Despite this, we believe that the technique as described semi-formally above is compelling. Adding support for tycon-specific typing contexts, implicit coercions and direct opcon calls can all be taken up by the community. Delving further into the question of when can two tycons with the same signature be substituted for one another, using a technique based on admissible relations, is also an avenue we wish to explore \cite{pfpl}.

%We note that is increasingly being deployed in contemporary languages (e.g. Scala) due to the perception that it enables good error messages, and to support semantics where  non-local type inference is undecidable. Permitting extensibility together with non-local type inference is an open problem.

Our work has some similarities to work on maintaining type abstraction using a form of effect system \cite{CraryHarperDreyer2002}, and on translating modules to System F \cite{conf/tldi/RossbergRD10}. Unlike modules, tycons permit the representation types to be computed, use a separate IL, and permit operations whose types cannot simply be written as functions. %Both examples we used here would be quite difficult to fully embed using only modules.

%\section{Conclusion}\label{conclusion}

%\section{Older Stuff}
%
%When only a weak embedding is possible, the solution is less clear. 
%
%
%
%
%even in the rare situation where a common framework has been used or the dialects are implemented within the same compiler, as these mechanisms do not guarantee that combining separately sound dialects will be sound.  Avoiding dialect composition is difficult, because interactions between components of an application written in different dialects leads to precisely the situation in item 3 above.% We will consider this further in our discussion of related work in Sec. \ref{related-work}.
%
%\paragraph{Language-Integrated Approaches}
%
%\paragraph{Active Type Constructors}
%
%\paragraph{Contributions}
%
% For example, the  fragment providing binary product types might define forms like $\texttt{(\textit{e}}\texttt{, \textit{e}}\texttt{)}$, $\texttt{$e$}\texttt{.l}$ and $\texttt{$e$}\texttt{.r}$, the term constructors $\FF{pair}$ and $\FF{prj}$ and the type constructor $\fvar{prod}$. A \emph{term} is constructed by applying a term constructor to an \emph{index} and zero or more \emph{arguments}. For example, \emph{expressions}, $e$, which are one \emph{sort} of term, are written $\FF{pair}[\FF{triv}](e_1; e_2)$, $\FF{prj}[\FF{left}](e)$ and $\FF{prj}[\FF{right}](e)$. Expressions are indexed by \emph{static terms}, $\sigma$. These are often simply assumed to be part of the ambient metatheory, but we will specify static term constructors explicitly. We omit trivial indices and empty argument lists for terms other than expressions for concision. 
%
%Types arise by applying a type constructor to a \emph{type index}, also a static term. For example, the type constructor $\fvar{prod}$ is indexed by a static pair of types (for $n$-ary products, the index would instead be a static list of types).  The sort where types are defined is commonly referred to as the \emph{type-level language}, but because types can appear in term and type indices, it is useful for us to include them directly in the static language. We might specify a trivially indexed static term constructor corresponding to each type constructor where the type index is the term argument, writing $\FF{prod}(\FF{spair}(\sigma_1; \sigma_2))$, or we might define a single static term constructor $\FF{ty}$ indexed by a type constructor, writing $\FF{ty}[\fvar{prod}](\FF{spair}(\sigma_1; \sigma_2))$. The static and dynamic semantics can be specified in a standard way \cite{pfpl}. %The static semantics serve as a specification for the typechecker and the dynamic semantics serves as a specification for the first stage of a compiler, which we call translation. In some cases, the static and dynamic semantics are given simultaneously by giving translation to an internal language (e.g. the Harper-Stone semantics of Standard ML \cite{harper-stone}). % Note that in practice, the dynamic semantics of expressions is implemented by translation to a simpler \emph{internal language}, with terms $\iota$ and types $\tau$. 
%
%
%
%%Modern languages like ML provide general-purpose type constructors that admit strong and often satisfying embeddings of many language fragments, occupying what is widely seen as a ``sweet spot'' in the design space. 
%
%New fragments are generally introduced as distinct \emph{dialects} of an existing general-purpose language, constructed either as a fork, using tools like compiler generators, DSL frame\-works or  language workbenches, or directly within a particular compiler for the language, often activated by a flag or pragma. 
%These are not the sorts of problems usually faced by library providers. Well-designed languages preclude the  possibility of ``link-time'' conflicts between libraries and ensure that the semantics of one library  cannot be weakened by another by strictly enforcing abstraction barriers. For example, a module declaring an abstract type in ML can rely on its representation invariants no matter which other modules are in use, so clients can assume that they will operate robustly in combination without needing to attend to burdensome ``link-time'' proof obligations. 
%
%
%
%
% number of researchers have observed that the problem of combining fragments to form a language is made complicated by problems of \emph{ambiguity}. The simplest forms of ambiguities are naming ambiguities: two type constructors or term constructors may have the same name. Globally unique name generation mechanisms are known to avoid this problem. 
%
%
%Types arise by applying a type constructor to a \emph{type index}.  We will view both types and type indices as  static terms. Types arise via the static term constructor $\FF{ty}$, which is indexed by a type constructor and has a  single argument, the type index, e.g. $\FF{ty}[\fvar{prod}](\FF{spair}(\sigma_1; \sigma_2))$, where $\sigma_1$ and $\sigma_2$ are two other types and $\FF{spair}$ is the static term constructor for static pairs (its trivial index is omitted). In the concrete syntax, we might write this instead $\fvar{prod}\texttt{[}\sigma_1; \sigma_2\texttt{]}$ or using a stylized infix notation, $\sigma_1 \times \sigma_2$. 
%
%A fragment is often identified by the type constructor it is organized around, so the STLC is  called $\mathcal{L}\{\fvar{arrow}\}$, following the convention used by Harper \cite{pfpl} and others (but avoiding ``stylistic sugar'' for the sake of uniformity).
%
%G\"odel's $\mathbf{T}$ consists of the $\rightarrow$ fragment and the \verb|nat| fragment, defining natural numbers and a recursor that allows one to ``fold'' over a natural number. We might thus call it $\mathcal{L}\{\rightarrow \mathtt{nat}\}$. This language is  more powerful than the STLC because the STLC admits an embedding  into  $\mathbf{T}$ but the reverse is not true. Buoyed by this fact, we might go on by adding fragments that define sums, products and various forms of inductive types (e.g. lists), or perhaps a general mechanism for defining inductive types. Each fragment clearly increases the expressiveness of our language.
%
%
%If we consider the $\forall$ fragment, however, defining universal quantification over types (i.e. parametric polymorphism), we must take a moment to reflect. In $\mathcal{L}\{\rightarrow\,\forall\}$, studied variously by Girard as System $\mathbf{F}$ \cite{girard1971extension} and Reynolds as the polymorphic lambda calculus \cite{Reynolds94anintroduction}, it is known that sums, products, and inductive and co-inductive types can all be weakly defined. This means that we can \emph{translate} well-typed terms of a language like $\mathcal{L}\{\rightarrow\forall~\mathtt{nat} + \times\}$ to well-typed terms in $\mathcal{L}\{\rightarrow\forall\}$ in a manner that preserves their dynamic semantics. But Reynolds, in a remark that recalls the ``Turing tarpit'' of Perlis \cite{Perl82a}, reminds us that programming with the corresponding \emph{embedding} may be unwise \cite{Reynolds94anintroduction}: 
%\begin{quote}
%To say that any reasonable function can be expressed by some program is not to say that it can be expressed by the most reasonable program. It is clear that the language requires a novel programming style. Moreover, it is likely that certain important functions cannot be expressed by their most efficient algorithms.
%\end{quote}
%
%%Adding new fragments directly to a language can make statically reasoning about programs more precise, support a more natural programming style and endow the language with a more favorable cost semantics. The latter two points might be relevant even if a strong embedding (i.e. one where there is a semantics-preserving isomorphism between terms and types of the fragment and the language) can be found. 
%%
%%Consistent with this view, typed programming languages like ML expose, for example, both $n$-ary product types and record types, despite the fact that any product type is definable using a record type (or in the other direction, using a module to provide the field selection operator). The compiler, on the other hand, \emph{is} mainly concerned with implementing the dynamic semantics correctly, and indeed, compilers often use the fact that weak definability results are quite readily derivable to their advantage, translating the constructs of an external language (EL) to those of a much simpler internal language (IL), e.g. with only binary products, during or directly after typechecking\todo{cite Harper-Stone}. %Correctly implementing the dynamic semantics of the language becomes the primary concern during this phase, so a simple IL that decreases the number of cases needing consideration when implementing and verifying the compiler is a wise choice.
%
%%Having established why a minimal language like $\mathcal{L}\{\rightarrow\forall\}$, while suitable as an internal language, needs to be extended with additional fragments before it is suitable for use as a human-facing external language, 
%A stronger embedding requires constructing an isomorphism that preserves both static reasoning principles and dynamic behaviors, in a sense that we will make more precise as we go on. This is generally far more difficult to establish than a weak embedding, and fewer universality results can be applied. For example, the aforementioned embeddings into $\mathcal{L}\{\rightarrow\forall\}$ do not preserve type disequality and some other equational reasoning principles (we will see less subtle issues soon). Establishing an isomorphic embedding that is natural (as measured, to a first approximation, by the ratio of boilerplate code that must be manually generated in the embedding vs. in the fragment, and how likely one is to make a mistake when writing it) and that has a reasonable cost semantics is harder still.
%
%
%\paragraph{Contributions}
%Drawing from this approach, we aim to introduce a mechanism that makes it possible to declare new type and operator constructors and define, \emph{using type-level functions}, their static and dynamic semantics, the latter by translation to a fixed typed internal language. If a language fragment can be weakly defined in terms of the typed internal language and its statics are of a general form consistent with this protocol, a n isomorphic embedding in the external language can be constructed by simply introducing the necessary constructors and defining directly the necessary type assignment and translation logic. We call such an embedding an \emph{active embedding}.
%
%The semantics imposes various checks to ensure type safety and decidability of type assignment and equality. Moreover, an important class of lemmas that play a key role in  proofs that a library enables strong embeddings of a language fragment can be derived in a suitable ``closed world'' (that is, in the traditional manner where one reasons inductively over a \emph{finite} collection of constructors). The semantics guarantees that they will be \emph{conserved} in the  ``open world'' by enforcing an abstraction barrier at extension boundaries. This avoids the most fundamental semantic problems that can arise when composing dialects and justifies the inclusion of the mechanism inside the language.
%
%We will begin in Sec. \ref{core} by introducing a minimal calculus with this kind of {actively typed} semantics, @$\lambda$. The external language begins with only the $\rightarrow$ fragment and the typed internal language is a simple variant of PCF that permits only weak embeddings of even simple fragments like the $\mathtt{nat}$ fragment of G\"odel's $\mathbf{T}$ or ML-style $n$-ary products. We actively embed these into the external language to explain the mechanism and motivate the constraints that the semantics imposes. We then examine the semantics in greater detail and state and sketch the key points in the proofs of the metatheory in Sec. \ref{theory}. These two sections comprise the fundamental contributions of this work.
%
%While the core calculus addresses the issue of making a strong embedding possible, active embeddings into the core calculus are often impractical syntactically. To begin to address the issue of syntax, we develop in Sec. \ref{expanded-syntax} a flexible type-directed dispatch protocol for a more conventional {expanded syntax} that defers to the core calculus semantically. We then use the expanded syntax to discuss several more interesting examples in Sec. \ref{examples}: labeled sums and products, the latter with delegation and functional field update, and a variant of \verb|sprintf|.\todo{update this if something doesn't make it}
%
%While of surprising expressiveness given its minimality, @$\lambda$ itself is certainly not capable of admitting embeddings of the full variety of type system fragments enumerated earlier. Indeed, to guarantee safety and conservativity and because our intention in this paper is to cleanly expose the essence of this mechanism, our calculus makes perhaps too many assumptions about the type system being embedded. For example, it only admits fragments where the typing judgement looks essentially like the typing judgement of the $\rightarrow$ fragment, so new contexts cannot be introduced. We discuss this and some of the other reasons why a fragment may not be satisfyingly embeddable in Sec. \ref{limitations}. These limitations do not appear fundamental, so we also outline directions for future work. We conclude by considering related work in Sec. \ref{related-work}. 
%
%\section{Core @$\lambda$}\label{core}
%%\newcommand{\pimport}[2]{\keyw{import}~#1~\keyw{in}~#2}
%%
%%\newcommand{\es}{\overline{e}}
%%\newcommand{\elet}[3]{\keyw{let}(#1; #2.#3)}
%%\newcommand{\eslet}[3]{\keyw{slet}[#1](#2.#3)}
%%\newcommand{\easc}[2]{\keyw{asc}[#1](#2)}
%%\newcommand{\elamrec}[3]{\keyw{lamrec}(#1,#2.#3)}
%%\newcommand{\elit}[2]{\keyw{lit}[#1](#2)}
%%\newcommand{\etarg}[3]{\keyw{targ}[#1](#2; #3)}
%%\newcommand{\ebranch}[3]{\keyw{branch}[#1](#2; #3)}i
%%
%%\newcommand{\qitm}{\grave\iota}
%%\newcommand{\qs}{\overline{\qitm}}
%%\newcommand{\qlet}[3]{\keyw{let}(#1; #2.#3)}
%%\newcommand{\qasc}[2]{\keyw{asc}[#1](#2)}
%%\newcommand{\qlamrec}[3]{\keyw{lamrec}(#1,#2.#3)}
%%\newcommand{\qlit}[2]{\keyw{lit}[#1](#2)}
%%\newcommand{\qtarg}[3]{\keyw{targ}[#1](#2; #3)}
%%\newcommand{\qap}[2]{\keyw{ap}(#1; #2)}
%%\newcommand{\qstr}[1]{\keyw{str}[#1]}
%%\newcommand{\qstrrec}[4]{\keyw{strrec}(#1; #2; {#3}.#4)}
%%\newcommand{\qpair}[2]{\keyw{pair}(#1; #2)}
%%\newcommand{\qletpair}[4]{\keyw{letpair}(#1; {#2}, {#3}.#4)}
%%\newcommand{\qnil}{\keyw{nil}}
%%\newcommand{\qcons}[2]{\keyw{cons}(#1; #2)}
%%\newcommand{\qlistrec}[6]{\keyw{listrec}(#1; #2; {#3}, {#4}, {#5}.#6)}
%%\newcommand{\qderivate}[3]{\keyw{derivate}[#1; #2; #3]}
%%\newcommand{\quq}[1]{{\lhd}[#1]}
%%
%%\newcommand{\hitm}{\hat{\iota}}
%%\newcommand{\hs}{\overline{\hitm}}
%%\newcommand{\hlet}[3]{\keyw{let}(#1; #2.#3)}
%%\newcommand{\hasc}[2]{\keyw{asc}[#1](#2)}
%%\newcommand{\hlamrec}[3]{\keyw{lamrec}(#1,#2.#3)}
%%\newcommand{\hlit}[2]{\keyw{lit}[#1](#2)}
%%\newcommand{\htarg}[3]{\keyw{targ}[#1](#2; #3)}
%%\newcommand{\hbranch}[3]{\keyw{branch}[#1](#2; #3)}
%%\newcommand{\hap}[2]{\keyw{ap}(#1; #2)}
%%\newcommand{\hstr}[1]{\keyw{str}[#1]}
%%\newcommand{\hstrrec}[4]{\keyw{strrec}(#1; #2; {#3}.#4)}
%%\newcommand{\htriv}{\keyw{triv}}
%%\newcommand{\hpair}[2]{\keyw{pair}(#1; #2)}
%%\newcommand{\hletpair}[4]{\keyw{letpair}(#1; {#2}, {#3}.#4)}
%%\newcommand{\hinl}[2]{\keyw{inl}[#1](#2)}
%%\newcommand{\hinr}[2]{\keyw{inr}[#1](#2)}
%%\newcommand{\hcase}[5]{\keyw{case}(#1; #2.#3; #4.#5)}
%%\newcommand{\hnil}{\keyw{nil}}
%%\newcommand{\hcons}[2]{\keyw{cons}(#1; #2)}
%%\newcommand{\hlistrec}[6]{\keyw{listrec}(#1; #2; {#3}, {#4}, {#5}.#6)}
%%\newcommand{\hderivate}[3]{\keyw{derivate}[#1; #2; #3]}
%%\newcommand{\huq}[1]{{\lhd}[#1]}
%%
%%\newcommand{\itm}{\iota}
%%\newcommand{\is}{\overline{itm}}
%%\newcommand{\ilet}[3]{\keyw{let}(#1; #2.#3)}
%%\newcommand{\iasc}[2]{\keyw{asc}[#1](#2)}
%%\newcommand{\ilamrec}[3]{\keyw{lamrec}(#1,#2.#3)}
%%\newcommand{\iap}[2]{\keyw{ap}(#1; #2)}
%%\newcommand{\istr}[1]{\keyw{str}[#1]}
%%\newcommand{\istrrec}[4]{\keyw{strrec}(#1; #2; {#3}.#4)}
%%\newcommand{\itriv}{\keyw{triv}}
%%\newcommand{\ipair}[2]{\keyw{pair}(#1; #2)}
%%\newcommand{\iletpair}[4]{\keyw{letpair}(#1; {#2}, {#3}.#4)}
%%\newcommand{\iinl}[2]{\keyw{inl}[#1](#2)}
%%\newcommand{\iinr}[2]{\keyw{inr}[#1](#2)}
%%\newcommand{\icase}[5]{\keyw{case}(#1; #2.#3; #4.#5)}
%%\newcommand{\inil}{\keyw{nil}}
%%\newcommand{\icons}[2]{\keyw{cons}(#1; #2)}
%%\newcommand{\ilistrec}[6]{\keyw{listrec}(#1; #2; {#3}, {#4}, {#5}.#6)}
%%
%%%\newcommand{\scolor}[1]{{\color{blue} #1}}
%%\newcommand{\st}{\sigma}
%%\newcommand{\stx}[1]{\st_\text{#1}}
%%\newcommand{\sval}{\hat\sigma}
%%\newcommand{\svalx}[1]{\sval_\text{#1}}
%%\newcommand{\sx}[1]{{\bf #1}}
%%\newcommand{\slam}[3]{\lambda[#1](\sx{#2}.#3)}
%%\newcommand{\sap}[2]{\keyw{ap}(#1; #2)}
%%\newcommand{\striv}{\keyw{triv}}
%%\newcommand{\sinl}[2]{\keyw{inl}[#1](#2)}
%%\newcommand{\sinr}[2]{\keyw{inr}[#1](#2)}
%%\newcommand{\scase}[5]{\keyw{case}(#1; \sx{#2}.#3; \sx{#4}.#5)}
%%\newcommand{\sfold}[3]{\keyw{fold}[#1.#2](#3)}
%%\newcommand{\srec}[4]{\keyw{rec}[#1](#2; \sx{#3}.#4)}
%%\newcommand{\sstr}[1]{\keyw{str}[#1]}
%%\newcommand{\sstrrec}[4]{\keyw{strrec}(#1; #2; \sx{#3}.#4)}
%%\newcommand{\spair}[2]{\keyw{pair}(#1; #2)}
%%\newcommand{\sletpair}[4]{\keyw{letpair}(#1; \sx{#2}, \sx{#3}.#4)}
%%\newcommand{\snil}{\keyw{nil}}
%%\newcommand{\scons}[2]{\keyw{cons}(#1; #2)}
%%\newcommand{\slistrec}[6]{\keyw{listrec}(#1; #2; \sx{#3}, \sx{#4}, \sx{#5}.#6)}
%%\newcommand{\sty}[3]{\keyw{ty}[\fvar{#1}](#2.#3)}
%%\newcommand{\selfvar}[1]{\textrm{#1}}
%%\newcommand{\m}{\selfvar{m}}
%%\newcommand{\stycase}[5]{\keyw{tycase}[\fvar{#1}](#2; \sx{#3}.#4; #5)}
%%\newcommand{\sarg}[1]{\keyw{arg}[#1]}
%%%\newcommand{\sanaarg}[5]{\keyw{ana}(#1; #2; \sx{#3},\sx{#4}.#5)}
%%%\newcommand{\ssynarg}[3]{\keyw{syn}(#1; \sx{#2}.#3)}
%%\newcommand{\sana}[3]{{\color{gray}\keyw{ana}[#1](#2; #3)}}
%%\newcommand{\ssyn}[2]{{\color{gray}\keyw{syn}[#1](#2)}}
%%\newcommand{\sraise}{\keyw{raise}}
%%\newcommand{\strtm}[1]{{\rhd}(#1)}
%%\newcommand{\strty}[1]{{\blacktriangleright}(#1)}
%%\newcommand{\saitm}[1]{{\hat\rhd}(#1)}
%%\newcommand{\saity}[1]{{\hat\blacktriangleright}(#1)}
%%\newcommand{\styeq}[4]{\keyw{tyeq}(#1; #2; #3; #4)}
%%
%%\newcommand{\kx}[1]{\kappa_\text{#1}}
%%\newcommand{\karrow}[2]{#1 \rightarrow #2}
%%\newcommand{\kstr}{\keyw{Str}}
%%\newcommand{\kprod}[2]{#1 \times #2}
%%\newcommand{\ksum}[2]{#1 + #2}
%%\newcommand{\kunit}{\keyw{1}}
%%\newcommand{\klist}[1]{\keyw{List}[#1]}
%%\newcommand{\kind}[2]{\keyw{ind}(#1.#2)}
%%\newcommand{\kty}{\keyw{Ty}}
%%\newcommand{\karg}{\keyw{Arg}}
%%\newcommand{\ktrtm}{\keyw{TrTm}}
%%\newcommand{\ktrty}{\keyw{TrTy}}
%%
%%\newcommand{\qity}{\grave\tau}
%%\newcommand{\qtparr}[2]{#1 \rightharpoonup #2}
%%\newcommand{\qtstr}{\keyw{str}}
%%\newcommand{\qtprod}[2]{#1 \times #2}
%%\newcommand{\qtlist}[1]{\keyw{list}[#1]}
%%\newcommand{\qtrepof}[1]{\keyw{rep}[#1]}
%%\newcommand{\qtuq}[1]{{\blacktriangleleft}[#1]}
%%
%%\newcommand{\hity}{\hat\tau}
%%\newcommand{\htparr}[2]{#1 \rightharpoonup #2}
%%\newcommand{\htstr}{\keyw{str}}
%%\newcommand{\htunit}{\keyw{1}}
%%\newcommand{\htprod}[2]{#1 \times #2}
%%\newcommand{\htsum}[2]{#1 + #2}
%%\newcommand{\htrec}[2]{\mu#1.#2}
%%\newcommand{\htlist}[1]{\keyw{list}[#1]}
%%\newcommand{\htrepof}[1]{\keyw{rep}[#1]}
%%\newcommand{\htuq}[1]{{\blacktriangleleft}[#1]}
%%
%%\newcommand{\ity}{\tau}
%%\newcommand{\itparr}[2]{#1 \rightharpoonup #2}
%%\newcommand{\itunit}{\keyw{1}}
%%\newcommand{\itstr}{\keyw{str}}
%%\newcommand{\itsum}[2]{#1 + #2}
%%\newcommand{\itrec}[2]{\mu#1.#2}
%%\newcommand{\itprod}[2]{#1 \times #2}
%%\newcommand{\itlist}[1]{\keyw{list}[#1]}
%%
%%\newcommand{\esyn}[5]{#1 \vdash_{#2} #3 \Rightarrow #4 \leadsto #5}
%%\newcommand{\esynX}[3]{\esyn{\Upsilon}{\Phi}{#1}{#2}{#3}}
%%
%%\newcommand{\eana}[5]{#1 \vdash_{#2} #3 \Leftarrow #4 \leadsto #5}
%%\newcommand{\eanaX}[3]{\eana{\Upsilon}{\Phi}{#1}{#2}{#3}}
%%
%%\newcommand{\etyerr}[3]{#1 \vdash_{#2} #3~\mathtt{tyerr}}
%%\newcommand{\etyerrX}[1]{\etyerr{\Upsilon}{\Phi}{#1}}
%%
%%\newcommand{\efragerr}[3]{#1 \vdash_{#2} #3~\mathtt{ferr}}
%%\newcommand{\efragerrX}[1]{\efragerr{\Upsilon}{\Phi}{#1}}
%%
%%\newcommand{\sevalCtx}{\mathcal{C}}
%%\newcommand{\sevalCtxNone}{\cdot}
%%\newcommand{\seval}[3]{#1 \Downarrow_{#2} #3}
%%\newcommand{\sevalC}[2]{\seval{#1}{\sevalCtx}{#2}}
%%\newcommand{\sevalN}[2]{\seval{#1}{\sevalCtxNone}{#2}}
%%
%%\newcommand{\tctxbind}[2]{#1 \Rightarrow #2}
%%
%%\newcommand{\kctx}{\Psi}
%%\newcommand{\skind}[4]{#1 \vdash_{#2} #3 : #4}
%%\newcommand{\skindX}[2]{\skind{\kctx}{\Phi}{#1}{#2}}
%%\newcommand{\args}{\mathtt{args}(\es)=\svalx{args};\Xi}
%%
%%\begin{figure*}[t]
%%\small
%%\hspace{-5px}
%%$
%%\begin{array}[t]{rlcl}
%%\mvlbl{programs} &
%%p & ::= &\pimport{\Phi}{e}\\
%%\mvlbl{tycon defs} & 
%%\Phi & ::= & \keyw{tycon}~\fvar{tc}[\kappa]~ \{\keyw{rep}=\st~\keyw{in}~\omega\} ~|~ \Phi; \Phi\\
%%\mvlbl{op defs} & \omega & ::= & \keyw{analit}[\kappa]=\st; \keyw{syntarg}[\kappa]=\st; \keyw{anabranch}[\kappa]=\st\\
%%\end{array}
%%$
%%\\
%%~\\
%%%\begin{tabular}{ll}
%%%$\begin{array}[t]{l}
%%%\mvlbl{static terms}\\
%%%\st ::= \sx{x}\\
%%%\slam{\kappa}{x}{\st}\\
%%%\sap{\st}{\st}\\
%%%...\\
%%%\spair{\st}{\st}\\
%%%\sletpair{\st}{x}{y}{\st}\\
%%%\snil\\
%%%\scons{\st}{\st}\\
%%%\slistrec{\st}{\st}{x}{y}{z}{\st}\\
%%%\sstr{s}\\
%%%\sstrrec{\st}{\st}{x}{\st}\\
%%%\sty{tc}{\st}\\
%%%\stycase{tc}{\st}{x}{\st}{\st}\\
%%%\sarg{n}\\
%%%\sanaarg{\st}{\st}{x}{y}{\st}\\
%%%\ssynarg{\st}{x}{\st}\\
%%%\strtm{\qitm}\\
%%%\strty{\qity}\\
%%%\styeq{\st}{\st}{\st}{\st}\\
%%%\sraise\\
%%%\sval\\
%%%\end{array}$~$\begin{array}[t]{l}
%%%\mvlbl{static values}\\
%%%\sval ::= \\
%%%\slam{\kappa}{x}{\st}\\
%%%...\\
%%%\spair{\sval}{\sval}\\
%%%\snil\\
%%%\scons{\sval}{\sval}\\
%%%\sstr{s}\\
%%%\sty{tc}{\sval}\\
%%%\sarg{n}\\
%%%\saitm{\hitm}\\
%%%\saity{\hity}\end{array}$
%%%$\begin{array}[t]{l}
%%%\\
%%%\mvlbl{tycon names}\\
%%%\fvar{tc}\\
%%%%\text{variables}\\
%%%%f, x, y, z\\
%%%%\text{static variables}\\
%%%%x, y, z\\
%%%\\
%%%\\
%%%\mvlbl{strings}\\
%%%s\\
%%%\text{nats}~~
%%%n
%%%\end{array}
%%%$
%%%$\begin{array}[t]{l}
%%%\mvlbl{quoted internal types}~~ \qity ::= \\
%%%%\begin{array}[t]{lcl}
%%%%\qity & ::= & 
%%%\qtparr{\qity}{\qity}\pipe
%%%\qtprod{\qity}{\qity}\pipe
%%%\qtlist{\qity}\pipe
%%%\qtstr%\\ & \pipe & 
%%%\pipe \qtrepof{\st} \pipe
%%%\qtuq{\st}%\end{array}\\
%%%\\
%%%\mvlbl{abstract internal types}~~\hity ::= \\
%%%%\begin{array}[t]{lcl}
%%%% & ::= & 
%%%\htparr{\hity}{\hity}\pipe
%%%\htprod{\hity}{\hity}\pipe
%%%1 \pipe + \pipe \mu \pipe 
%%%%\htlist{\hity}\pipe
%%%%\htstr\pipe%\\ & \pipe & 
%%%\htrepof{\sval}
%%%\\
%%%\mvlbl{internal types}~~\ity ::= \\
%%%%\begin{array}[t]{lcl}
%%%%\ity & ::= & 
%%%\itparr{\ity}{\ity}\pipe
%%%\itprod{\ity}{\ity}\pipe
%%%\itlist{\ity}\pipe
%%%\itstr%\end{array}\\
%%%%\end{array}
%%%%\begin{array}[t]{lcll}
%%%%\Delta & ::= & \cdot \pipe \Delta, \sx{x}:\kappa & \mvlbl{kinding ctx}\\
%%%%\Upsilon & ::= & \cdot \pipe \Upsilon, x\Rightarrow\sval & \mvlbl{typing ctx}\\
%%%%\mathcal{C} & ::= & \cdot \pipe \es;\Upsilon;\Phi & \mvlbl{current op}\\
%%%%\mvlbl{abstract internal typing context}~\Psi ::= ...\\
%%%%\Xi & ::= & \cdot \pipe \Xi, \fvar{tc} & \mvlbl{tycon list}\\
%%%%\es & ::= & \cdot \pipe e; \es & \mvlbl{term lists}\\
%%%%\qs & ::= & \cdot \pipe \qitm; \qs\\
%%%%\hs & ::= & \cdot \pipe \hitm; \hs\\
%%%%\end{array}
%%%\end{array}$\\
%%%\end{tabular}~\\
%%\begin{tabular}{>{$}l<{$}>{$}l<{$}|>{$}l<{$}|>{$}l<{$}>{$}l<{$}|>{$}l<{$}>{$}l<{$}}
%%\multicolumn{2}{l|}{\mvlbl{kinds \& static terms}} & \mvlbl{external terms} & \multicolumn{2}{l|}{\mvlbl{translational types \& terms}} & \multicolumn{2}{l}{\mvlbl{internal types \& terms}}\\
%%\kappa ::= & \st ::= & e ::= & \hity ::= & \hitm ::= & \ity ::= & \itm ::=\\
%%k & \sx{x} & x & t & x & t & x\\
%%\kty & \sty{tc}{\m}{\st} & \elet{e}{x}{e} & & \hlet{\hitm}{x}{\hitm} & & \ilet{\itm}{x}{\itm}\\
%%& \m & \\
%%& \stycase{tc}{\st}{x}{\st}{\st} & \easc{\st}{e} & & \hasc{\hity}{\hitm} & & \iasc{\ity}{\itm}\\
%%& \styeq{\st}{\st}{\st}{\st} & \elit{\st}{\es} & & \hlit{\st}{\hs}\\
%%\karrow{\kappa}{\kappa} & \slam{\kappa}{x}{\st} & \etarg{\st}{e}{\es} & & \htarg{\st}{h}{\hs} & \\
%% & \sap{\st}{\st} & \ebranch{\st}{e}{\es} & & \hbranch{\st}{h}{\hs}\\
%%\kunit & \striv & \elamrec{f}{x}{e} & \htparr{\hity}{\hity} & \hlamrec{f}{x}{\hitm} & \itparr{\ity}{\ity} & \ilamrec{f}{x}{\itm}\\
%%\kprod{\kappa}{\kappa} & \spair{\kappa}{\kappa} & & & \hap{\hitm}{\hitm} & & \iap{\itm}{\itm}\\
%%& \sletpair{\st}{x}{y}{\st} & & \htunit & \htriv & \itunit & \itriv\\
%%\ksum{\kappa}{\kappa} & \sinl{\kappa}{\st} & & \htprod{\hity}{\hity} & \hpair{\hitm}{\hitm} & \itprod{\ity}{\ity} & \ipair{\itm}{\itm}\\
%%& \sinr{\kappa}{\st} & & & \hletpair{\hitm}{x}{y}{\hitm} & & \iletpair{\itm}{x}{y}{\itm}\\
%%& \scase{\st}{x}{\st}{x}{\st} & & \htsum{\hity}{\hity} & \hinl{\hity}{\hitm} & \itsum{\ity}{\ity} & \iinl{\ity}{\itm}\\
%%\kind{k}{\kappa} & \sfold{k}{\kappa}{\st} & & & \hinr{\hity}{\hitm} & & \iinr{\ity}{\itm}\\
%%& \srec{\kappa}{\st}{x}{\st} & & & \hcase{\hitm}{x}{\hitm}{x}{\hitm} & & \icase{\itm}{x}{\itm}{x}{\itm}\\
%%\ktrty & \strty{\hity} & & \htrec{t}{\hity}\\
%%\ktrtm & \strtm{\hitm}\\
%%& \sana{n}{\st}{\st}\\
%%& \ssyn{n}{\st}\\
%%& \sraise & ~& & &2
%%\end{tabular}\\
%%$\begin{array}[t]{l}
%%\mvlbl{external terms}\\
%%e ::= x\\
%%\elet{e}{x}{e}\\
%%\easc{\st}{e}\\
%%\elit{\st}{\es}\\
%%\etarg{\st}{e}{\es}\\
%%\ebranch{\st}{e}{\es}\\
%%\elamrec{f}{x}{e}\\
%%%\qderivate{n}{\sval}{\hitm}\\
%%\end{array}$~
%%$\begin{array}[t]{l}
%%%~\\
%%\mvlbl{static terms}\\
%%\st ::= \sx{x} \\
%%\sty{tc}{t}{\st} \\
%% \\
%%\slam{\kappa}{x}{\st} \\ 
%%\sap{\st}{\st} \\
%%\striv\\
%%\spair{\st}{\st}\\
%%\sletpair{\st}{x}{y}{\st}\\
%%\sinl{\kappa}{\st}\\
%%\sinr{\kappa}{\st}\\
%%\scase{\st}{x}{\st}{x}{\st}\\
%%~\\
%%~\\
%%  \strtm{\hitm} \\
%%  \strty{\hity}\\
%%\sana{n}{\st}{\st} \\
%% \ssyn{n}{\st}  \\
%%% & & \pipe & \styeq{\st}{\st}{\st}{\st}\pipe\sraise\\
%%%\mvlbl{kinds} & \kappa & ::= & k \pipe \karrow{\kappa}{\kappa} \pipe \kunit \pipe \kprod{\kappa}{\kappa} \pipe \ksum{\kappa}{\kappa} \pipe \kind{k}{\kappa} \\
%%% & & \pipe & \kty \pipe \ktrtm \pipe \ktrty\\
%%%~\\
%%\end{array}$~
%%$\begin{array}[t]{l}
%%\mvlbl{kinds}\\
%%\kappa ::= k\\
%%\kty\\
%%~\\
%%\karrow{\kappa}{\kappa}\\
%%~\\
%%\kunit\\
%%\kprod{\kappa}{\kappa}\\
%%~\\
%%\ksum{\kappa}{\kappa}\\
%%~\\
%%~\\
%%\kind{k}{\kappa}\\
%%~\\
%%\ktrtm\\
%%\ktrty
%%\end{array}$~
%%$\begin{array}[t]{l}
%%\mvlbl{translational terms}\\
%%\hitm ::= x\\
%%\hlet{\hitm}{x}{\hitm}\\
%%\hasc{\hity}{\hitm}\\
%%\hlit{\sval}{\hs}\\
%%\htarg{\sval}{\hitm}{\hs}\\
%%\hbranch{\st}{\hitm}{\hs}\\
%%\hlamrec{f}{x}{\hitm}\\
%%\hap{\hitm}{\hitm}\\
%%\hpair{\hitm}{\hitm}\\
%%\hletpair{\hitm}{x}{y}{\hitm}\\
%%\hnil\\
%%\hcons{\hitm}{\hitm}\\
%%%\hlistrec{\hitm}{\hitm}{x}{y}{z}{\hitm}\\
%%\hstr{s}\\
%%\hstrrec{\hitm}{\hitm}{x}{\hitm}\\
%%\quq{\st}\\
%%\hderivate{n}{\sval}{\hitm}\\
%%\end{array}$~
%%$\begin{array}[t]{l}
%%\mvlbl{translational types}\\
%%%\mvlbl{types}\\
%%\hity ::= t\\
%%~\\
%%~\\
%%~\\
%%~\\
%%~\\
%%arrow\\
%%~\\
%%1\\
%%prod\\
%%~\\
%%sum\\
%%~\\
%%~\\
%%rec\\
%%~\\
%%uq\\
%%repoft
%%\end{array}$~
%%$\begin{array}[t]{l}
%%\mvlbl{internal terms}\\
%%\dot{\iota} ::= x\\
%%  \ilet{\itm}{x}{\itm}\\
%%\iasc{\ity}{\itm}\\
%%~\\
%%~\\
%%~\vspace{3px}\\
%%\ilamrec{f}{x}{\itm}\\
%%\iap{\itm}{\itm}\\
%%\ipair{\itm}{\itm}\\
%%\iletpair{\itm}{x}{y}{\itm}\\
%%\inil\\
%%\icons{\itm}{\itm}\\
%%\ilistrec{\itm}{\itm}{x}{y}{z}{\itm}\\
%%\istr{s}\\
%%\istrrec{\itm}{\itm}{x}{\itm}
%%\end{array}$~
%%$\begin{array}[t]{l}
%%\mvlbl{internal types}\\
%%\hity ::= t\\
%%~\\
%%~\\
%%~\\
%%~\\
%%~\\
%%arrow\\
%%~\\
%%1\\
%%prod\\
%%~\\
%%sum\\
%%~\\
%%~\\
%%rec\\
%%\end{array}$
%%\caption{Test}
%%\end{figure*}
%%\begin{figure*}
%%\small
%%\fbox{$\esynX{e}{\sval}{\hitm}$}~
%%\fbox{$\eanaX{e}{\sval}{\hitm}$}~
%%%\fbox{$\etyerrX{e}$}~
%%%\fbox{$\efragerrX{e}$}
%%\begin{mathpar}
%%\inferrule[syn-var]{
%%	\tctxbind{x}{\sval} \in \Upsilon
%%}{	
%%	\esynX{x}{\sval}{x}
%%}
%%
%%%\inferrule[tyerr-var]{
%%%	\tctxbind{x}{\sval} \notin \Upsilon
%%%}{
%%%	\etyerrX{x}
%%%}
%%%
%%\inferrule[syn-let]{
%%   \esynX{e_1}{\sval_1}{\hitm_1}\\
%%   \esyn{\Upsilon, \tctxbind{x}{\sval_1}}{\Phi}{e_2}{\sval_2}{\hitm_2}
%%}{
%%	\esynX{\elet{e_1}{x}{e_2}}{\sval_2}{\hlet{\hitm_1}{x}{\hitm_2}}
%%}
%%
%%\inferrule[ana-let]{
%%   \esynX{e_1}{\sval_1}{\hitm_1}\\
%%   \eana{\Upsilon, \tctxbind{x}{\sval_1}}{\Phi}{e_2}{\sval_2}{\hitm_2}
%%}{
%%	\eanaX{\elet{e_1}{x}{e_2}}{\sval_2}{\hlet{\hitm_1}{x}{\hitm_2}}
%%}
%%
%%\inferrule[syn-slet]{
%%	\skind{\cdot}{\Phi}{\st}{\kappa}\\
%%	\esynX{[\st/x]e}{\sval}{\hitm}
%%}{
%%	\esynX{\eslet{\st}{x}{e}}{\sval}{\hitm}
%%}
%%
%%\inferrule[ana-slet]{
%%	\skind{\cdot}{\Phi}{\st}{\kappa}\\
%%	\eanaX{[\st/x]e}{\sval}{\hitm}
%%}{
%%	\eanaX{\eslet{\st}{x}{e}}{\sval}{\hitm}
%%}
%%
%%\inferrule[syn-asc]{
%%	\skind{\cdot}{\Phi}{\st}{\kty}\\
%%	\sevalN{\st}{\sval}\\
%%	\eanaX{e}{\sval}{\hitm}
%%}{
%%	\esynX{\easc{\st}{e}}{\sval}{\hitm}
%%}
%%
%%\inferrule[ana-lamrec]{
%%	\Phi_0 \in \Phi\\
%%	\eana{\Upsilon, \tctxbind{f}{\sty{parrow}{t}{\spair{\sval_1}{\sval_2}}}, \tctxbind{x}{\sval_1}}{\Phi}{e}{\sval_2}{\hitm}
%%}{
%%	\eanaX{\elamrec{f}{x}{e}}{\sty{parrow}{t}{\spair{\sval_1}{\sval_2}}}{\hlamrec{f}{x}{\hitm}}
%%}
%%
%%\inferrule[ana-lit]{
%%  \fvar{tc}[\kx{tyidx}] \{\keyw{analit}[\kx{tmidx}]=\stx{analit}; \cdots\} \in \Phi\\\\
%%  \skind{\cdot}{\Phi}{\svalx{tmidx}}{\kx{tmidx}}\\
%%  \args\\\\
%%  \seval{\stx{analit}~\sty{tc}{t}{\svalx{tyidx}}~\svalx{tmidx}~\svalx{args}}{\Upsilon;\Phi;\Delta}{\saitm{\hitm}}\\\\
%%  check
%%}{
%%  \eanaX{\elit{\svalx{tmidx}}{\es}}{\sty{tc}{t}{\svalx{tyidx}}}{\hitm}
%%}
%%\end{mathpar}
%%\caption{Abstract Active Typechecking and Translation}
%%\end{figure*}
%%\inferrule[import]{
%%	\vdash \Phi\\
%%	\Upsilon \vdash_\Phi e \Rightarrow \sty{tc}{\sval} \leadsto \iota
%%}{
%%	\keyw{import}~\Phi~\keyw{in}~e : \sty{tc}{\sval} \leadsto \iota
%%}
%
%%
%%\begin{figure}[t]
%%\begin{lstlisting}
%%tycon Nat of 1 with  
%%    schema $\lambda$idx:1.$\blacktriangledown$($\mathbb{Z}$) 
%%    opcon Z of 1 ($\lambda$idx:1.$\lambda$args:list[Elab].is_empty args $\llbracket$$\triangledown$(0) as Nat[()]$\rrbracket$)
%%    opcon S of 1 ($\lambda$idx:1.$\lambda$args:list[Elab].pop_final args $\lambda$x:ITm.$\lambda$ty:$\star$. 
%%      checktype ty Nat[()] $\llbracket$$\triangledown$($\vartriangle$(x)+1) as Nat[()]$\rrbracket$)
%%    opcon Rec of 1 ($\lambda$idx:1.$\lambda$args:list[Elab].
%%      pop args $\lambda$x1:ITm.$\lambda$ty1:$\star$.$\lambda$args':list[Elab].
%%      pop args' $\lambda$x2:ITm.$\lambda$ty2:$\star$.$\lambda$args'':list[Elab].
%%      pop_final args'' $\lambda$x3:ITm.$\lambda$ty3:$\star$.
%%      check_type t1 Nat[()] (
%%      check_type t3 Arrow[(Nat[()], Arrow[(t2, t2)])] 
%%      	$\llbracket$$\triangledown$((fix f:$\mathbb{Z} \rightarrow$ rep(t2) is $\lambda$x:$\mathbb{Z}$.
%%	        if x = 0 then $\vartriangle$(x2) else $\vartriangle$(x3) (x - 1) (f (x - 1))) $\vartriangle$(x1)) as t2$\rrbracket$))
%%end
%%\end{lstlisting}
%%\caption{An implementation of primitive natural numbers as internal integers in @$\lambda$.}
%%\label{nat-atlam}
%%\end{figure}
%
%The syntax of Core @$\lambda$ is given in Fig. \ref{grammar}. An example of a program defining type and operator constructors that admit an active embedding of G\"odel's \textbf{T} into @$\lambda$ is given in Fig. \ref{nat}. We will discuss its semantics and how precisely that embedding, seen being used starting on line 16, works as we go on. Natural numbers can, of course, be strongly embedded in existing languages, with a similar usage and asymptotic performance profile (up to function call overhead as an abstract type, for example). We will provide more sophisticated examples where this is less feasible later on (and note that type abstraction itself ..... in Sec. \ref{limitations}).\todo{if i don't have time/space, take this out} %It consists of:
%%a type constructor declaration, $\fvar{Nat}$, indexed trivially, together with three operator constructors, also all indexed trivially, that implement the standard introductory forms for natural numbers as well as the recursor operator (as in G\"odel's T \cite{pfpl}). Following the type constructor declaration, we apply $\fvar{Nat}$ with the trivial index, $\tunit$, to form the type $\tvar{nat}$. Finally, we write an external term that uses the operators associated with $\fvar{nat}$ and the built-in constructor $\fvar{Parr}$, governing partial functions, to define an addition function and compute the addition of the natural numbers  two and two. We will introduce a more convenient concrete syntax in later portions of this thesis; for now we will restrict ourselves to the abstract syntax so that this example can directly aid in understanding the semantics.
%
%\subsection{Overview}\label{programs}
%%\begin{figure}[t]
%%\small
%%$$\begin{array}{rccl}	
%%\textbf{programs} & \rho & ::= & \pfam{\familyDf}{\progsort}
%%\\& &  \pipe & 
%%%\pipe \pdef{t}{\kappa}{\tau}{\progsort} 
%% e\\
%%		&	\theta	&	::= &	\tops{op}{\kappaidx}{i}{a}{\taut{def}} \pipe 
%%												\topp{\theta}{\theta}\\
%%\\
%%\textbf{external terms} 				&	e	&	::=	&	\evar{x} \pipe 
%%%														\efix{x}{\tau}{e} \pipe 
%%														%\elam{\evar{x}}{\tau}{e} \pipe 
%%														\eop{Tycon}{op}{
%%															\tauidx
%%														}{
%%  												    		\splat{e}{1}{n}
%%														} \\
%%									& 		&		& 	\\
%%
%%
%%\textbf{internal terms} 				& 	\iota	&	::=	&	\evar{x} \pipe 
%%												\ifix{\evar{x}}{\sigma}{\iota} \pipe
%%												\ilam{\evar{x}}{\sigma}{\iota} \pipe 
%%												\iapp{\iota_{1}}{\iota_{2}} 
%%\\\text{integers}&&\pipe&												
%%												\iintlit \pipe \iop{\iota_{1}}{\iota_{2}} \pipe \iIfEq{\iota_{1}}{\iota_{2}}{\dint}{\iota_{3}}{\iota_{4}} \\
%%\text{products}										& & \pipe & 
%%												\iunit \pipe
%%												\ipair{\iota_{1}}{\iota_{2}} \pipe 
%%												\ifst{\iota} \pipe
%%												\isnd{\iota} 
%%\\\text{sums}&&\pipe&												
%%												 \iinl{\sigma_2}{\iota_1} \pipe \iinr{\sigma_1}{\iota_2} \\&&\pipe& \icase{e}{x}{e_1}{x}{e_2} \\
%%												
%%%\text{deabstracted}& \iota & ::= & \mathcal{G}[\iota, \sigma]\\
%%\textbf{internal types}			&	\sigma	&	::=	&    \darrow{\sigma_1}{\sigma_2} \pipe \dint \pipe \dunit \pipe \dpair{\sigma_1}{\sigma_2} \pipe \dsum{\sigma_1}{\sigma_2}  
%%												
%%\\
%%\\
%%							
%%\hspace{-5pt}\textbf{type-level terms} 	& \tau 	& ::= 	& 	\tvar{t} \pipe 
%%														\tlam{t}{\kappa}{\tau} \pipe 
%%														\tapp{\tau_1}{\tau_2} \pipe \iintlit \pipe \iop{\tau_1}{\tau_2} \pipe \tlabel{label}
%%\\\text{lists}&&\pipe&											
%%														\tnil{\kappa} \pipe \tcons{\tau_1}{\tau_2} \pipe 
%%									                     \tfold{\tau_1}{\tau_2}{h}{t}{r}{\tau_3}
%%														\\
%%												
%%\text{products}	 			& 		& \pipe	& 	  \tunit \pipe 
%%														\tpair{\tau_{1}}{\tau_{2}} \pipe 
%%														\tfst{\tau} \pipe 
%%														\tsnd{\tau} 
%%														\\	
%%\text{sums}									&       & \pipe & \tinl{\kappa_2}{\tau_1} \pipe \tinr{\kappa_1}{\tau_2} \\&&\pipe& \tsumcase{\tau}{t}{\tau_1}{t}{\tau_2} \\
%%						\text{types} 						& 		& \pipe	& 	\ttypestd \\&&\pipe& \tfamcase{\tau}{Tycon}{x}{\tau_1}{\tau_2}\\
%%\text{equality}  & & \pipe & 					\tifeq{\tau_{1}}{\tau_{2}}{\kappa}{\tau_{3}}{\tau_{4}} 
%%														\\													%				& & \pipe & \tfamcase{\tau}{Fam}{x}{\tau_1}{\tau_2}\\
%%																								
%%\text{derivates} 				& 		 & 	\pipe	&	\tden{{\bar\iota}}{\tau} \pipe \tden{\ibar}{\tau}^{\checkmark} \pipe \ttypeof{\tau} \\% \tdencase{\tau}{x}{t}{\tau_1}{\tau_2}\\
%% %& & \pipe & 
%%%														\tdencase{\tau}{y}{x}{\tau_1}{\tau_2}
%%%														 \\
%%
%%\text{rep types}		&		&	\pipe	&	\titype{\bar \sigma} \\
%%% &  & \pipe & \tvalof{\tau_1}{\tau_2} \pipe \iup{\tau} \\
%%% 												\trepof{\tau} \pipe \dup{\tau}\\
%%\text{translational IL}	& \bar{\iota} & ::= & x \pipe \ifix{x}{\bar \sigma}{\bar \iota} 
%%	%\pipe \ilam{x}{\bar \sigma}{\bar \iota} \pipe \iapp{\bar \iota_1}{\bar \iota_2} 
%%	\pipe \cdots \pipe \itransof{\tau} \\						
%% & \bar{\sigma} & ::= & \darrow{\bar \sigma_1}{\bar \sigma_2} \pipe \cdots \pipe \dup{\tau} \pipe \trepof{\tau} \\
%%%\text{abstracted} & \sabs & ::= & \darrow{\sabs_1}{\sabs_2} \pipe \cdots \pipe \sabsrep{\tau}
%%											\\
%%\textbf{kinds} 					& \kappa	&	::=	&	\karrow{\kappa_1}{\kappa_2} \pipe \kint \pipe
%%											    \klabel \pipe
%%											    \klist{\kappa} \pipe
%%												\kunit \pipe 
%%												\kpair{\kappa_{1}}{\kappa_{2}} \\
%%												&&\pipe&
%%												\ksum{\kappa_1}{\kappa_2} \pipe
%%												\kTypeBlur \pipe \kDen \pipe 
%%												\kIType								
%%%\textbf{ops signature}			& \Theta	&	::=	&	\kOpEmpty \pipe \kOp{\Theta}{op}{\kappai}\\
%%%											 							&		&		&	\\
%%\end{array}$$
%%%\vspace{-10pt}
%%\caption{\small Syntax of Core \atlam. Here, $x$ ranges over external and internal language variables, $\tvar{t}$ ranges over type-level variables, $\fvar{Tycon}$ ranges over type constructor names, $\opvar{op}$ ranges over operator constructor names, $\iintlit$ ranges over integer literals, $\tlabel{label}$ ranges over label literals (see text) and $\oplus$ ranges over standard total binary operations (e.g. addition, comparison).
%%\label{grammar}}
%%\end{figure}
%%\begin{figure}[t]
%%\small
%%\begin{flalign}
%%\label{natfam}&\family{Nat}{\kunit}{\\
%%\label{z}&\quad\tops{z}{\kunit}{i}{a}{\tlam{i}{\kunit}{\tlam{a}{\klist{\kDen}}{
%%	\tapp{\tapp{\tvar{arity0}}{\tvar{a}}}{\\&\quad\quad
%%		\tden{0}{\ttype{Nat}{\tunit}}
%%	}}}};\\
%%\label{s}&\quad\tops{s}{\kunit}{i}{a}{\tlam{i}{\kunit}{\tlam{a}{\klist{\kDen}}{\tvar{arity1}~\tvar{a}~\tlam{d}{\kDen}{\\
%%& \quad\quad \tvar{ifeq}~\ttypeof{\tvar{d}}~\ttype{Nat}{\tunit}\\
%%& \quad\quad\quad \tden{\itransof{\tvar{d}} + 1}{\ttype{Nat}{\tunit}}
%%	}}}};\\
%%\label{rec}&\quad\tops{rec}{\kunit}{i}{args}{\tlam{i}{\kunit}{\tlam{a}{\klist{\kDen}}{\tvar{arity3}~\tvar{a}~\tlam{d1}{\kDen}{\tlam{d2}{\kDen}{\tlam{d3}{\kDen}{\\
%%&\quad\quad \tvar{ifeq}~\ttypeof{\tvar{d1}}~\ttype{Nat}{\tunit}\\
%%&\quad\quad {\sf let}~\tvar{t2} = \ttypeof{\tvar{d2}}~{\sf in}\\
%%&\quad\quad \tvar{ifeq}~\ttypeof{\tvar{d3}}~\ttype{Arrow}{(\ttype{Nat}{\tunit}, \ttype{Arrow}{(\tvar{t2}, \tvar{t2})})}\\
%%&\quad\quad\quad \tden{\iapp{(\ifix{f}{\darrow{\dint}{\trepof{\tvar{t2}}}}{\ilam{x}{\dint}{\\
%%		\label{lastop}&\quad\quad\quad\quad\quad \iIfEq{x}{0}{\dint}{\itransof{\tvar{d2}}}{\\
%%		&\quad\quad\quad\quad\quad\quad\iapp{\iapp{\itransof{\tvar{d3}}}{(x-1)}}{(\iapp{f}{(x-1)})}}}}\\
%%		&\quad\quad\quad\quad)}{\itransof{\tvar{d1}}}}{\tvar{t2}}
%%}}}
%%}}}
%%\\
%%&}{XXX}{i}{\tlam{i}{\kunit}{\titype{\dint}}};\\
%%\label{nattype}&{\sf let}~\tvar{nat} = \ttype{Nat}{\tunit}~{\sf in}\\
%%& \elet{two}{\eop{Nat}{s}{\tunit}{\eop{Nat}{s}{\tunit}{\eop{Nat}{z}{\tunit}{}}}}{\\
%%& \elet{plus}{
%%	\elam{x}{\tvar{nat}}{
%%		\elam{y}{\tvar{nat}}{
%%			\\&\quad \eop{Nat}{rec}{\tunit}{x; y; \elam{p}{\tvar{nat}}{\elam{r}{\tvar{nat}}{
%%				\eop{Nat}{s}{\tunit}{r}
%%			}}}
%%		}
%%	}
%%}{
%%& \\
%%& \eop{Arrow}{ap}{\tunit}{plus; \eop{Arrow}{ap}{\tunit}{two; two}}
%%}}
%%%{\\&\elet{two}{\tvar{nat}}{\eop{Nat}{s}{\tunit}{\eop{Nat}{s}{\tunit}{\eop{Nat}{z}{\tunit}{ }}}}{\eapp{\eapp{plus}{two}}{two}}}
%%%}
%%%\elam{x}{\tvar{nat}}{\elam{y}{\tvar{nat}}{\\
%%	%&\quad \eop{Nat}{rec}{\tunit}{x; y; \elam{p}{\tvar{nat}}{\elam{r}{\tvar{nat}}{
%%	%\eop{Nat}{s}{\tunit}{r}
%%	%}}}}
%%\end{flalign}
%%%\vspace{-15px}
%%\caption{\small An embedding of G\"odel's $\mathbf{T}$ in \atlam, used to define $plus$ and $two$ and calculate $plus~two~two$. Type constructors (line 1) and types (line 16) are discussed in Sec. \ref{types}. Schemas (line 1) and operator constructors (lines 2-14, used on lines 17-20) are discussed starting in Sec. \ref{opcons}. We use \textsf{let} to bind both external and type-level variables as needed for clarity of presentation. We will discuss further syntactic issues in Sec. \ref{expanded-syntax}.}
%%\label{nat}
%%\vspace{-10pt}
%%\end{figure}
%A \emph{program}, $\rho$, consists of a series of static declarations (in the core, only constructor declarations) followed by an external term, $e$. The syntax for external terms contains three forms: variables, $\lambda$ terms, and a form for invoking user-defined operators, which we will discuss below. Compiling a program consists of first \emph{kind checking} it then typechecking the external term and {translating} it to a term, $\iota$, in the {typed internal language}. These correspond to the premises of the \emph{central compilation judgement} $\pkcompiles{\rho}{\iota}$:
%\[
%\inferrule[p-compiles]{
%	\emptyset \vdash_{\fvalCtx_0} \rho\\
%%	\progOK{\emptyset}{\fvalCtx_0}{\rho}\\
%	\pcompiles{\fvalCtx_0}{\rho}{\iota}
%}{\pkcompiles{\rho}{\iota}}
%\]
%This anchors our exposition; we will describe how it is derived (i.e. how to write a compiler for @$\lambda$) in the following sections. 
%
%The key judgement in the calculus is the \emph{active typing judgement}  (Fig. \ref{att}, which we describe starting in Sec. \ref{opcons}). It relates an external term, $e$, to a {type}, $\tau$, called its \emph{type assignment}, and an internal term, $\iota$, called its \emph{translation}, under \emph{typing context} $\Upsilon$ and \emph{constructor context} $\Phi$: 
%\[\ecompilesX{e}{\tau}{\iota}\]
%
%The typing context $\Upsilon$ maps variables to types in essentially the conventional way (\cite{pfpl} contains the necessary background for this paper). The constructor context tracks user-defined type and operator constructors. % Weakening and exchange of the constructor context closely related to the issue of conservativity that we will return to. Each constructor is identified by name, so there is no analog to contraction.
%There is no separate operational semantics for the external language. Instead, the dynamic behavior of an external term is determined by its translation to the internal language, which has a more conventional operational semantics. This form of semantics can be seen as lifting into the language specification the first stage of a type-directed compiler like the TIL compiler for Standard ML \cite{tarditi+:til-OLD} and has some parallels to the Harper-Stone semantics for Standard ML, where external terms were also given meaning by elaboration from the EL to an IL \cite{Harper00atype-theoretic}. %
%
%In @$\lambda$, the internal language (IL) provides partial functions (via the generic fixpoint operator of Plotkin's PCF), simple product and sum types and a base type of integers (to make our example interesting and as a nod toward speed on contemporary machines). In practice, the internal language could be any typed  language with a specification for which type safety and decidability of typechecking have been satisfyingly determined. The internal type system serves as a ``floor'': guarantees that must hold for \emph{all} well-typed terms, independent of the constructor context (e.g. that out-of-bounds access to memory never occurs), must be maintained by the internal type system. User-defined constructors can (only) enforce invariants stronger than those the internal type system maintains. Performance is also ultimately limited by the internal language and downstream compilation stages that we do not here consider (safe compiler extension has been discussed in previous work, e.g. \cite{conf/pldi/TatlockL10}).
%
%\subsection{Type Constructors and Types}\label{types}
%\begin{figure}[t]
%\small
%$\fbox{\inferrule{}{\progOKX{\progsort}}}$
%~~~$\tvarCtx ::= \emptyctx \pipe \tvarCtxX{t}{\kappa}$
%~~~$\fvalCtx ::= \emptyctx \pipe \fvalCtxX{\fvalDf}$
%%$\fCtx ::= \Sigma_0	 \pipe \fvalCtxX$
%\begin{mathpar}
%\inferrule[k-tycon]{
%	\fvar{Tycon} \notin \text{dom}(\fvalCtx)\\
%	\kEq{\kappaidx}\\
%	\tKind{\tvarCtx}{\fvalCtx}{\taurep}{\karrow{\kappaidx}{\kIType}}\\
%	\opType{\tvarCtx}{\fvalCtxX{\fvalDf}}{\theta}{\Theta}\\
%	\progOK{\tvarCtx}{\fvalCtxX{\fvalDf}}{\rho}
%}{
%	\progOKX{\pfam{\familyDf}{\rho}}
%}
%
%%\inferrule[def-kinding]{
%%	\tKindX{\tau}{\kappa}\\
%%	\progOK{\tvarCtxX{t}{\kappa}}{\fvalCtx}{\rho}
%%}{
%%	\progOKX{\pdef{t}{\kappa}{\tau}{\rho}}
%%}
%%
%\inferrule[k-e-prog]{
%	\exprOK{\tvarCtx}{\fvalCtx}{e}
%}{
%	\progOKX{e}
%}
%\end{mathpar}
%$\fbox{$\opType{\tvarCtx}{\fvalCtx}{\theta}{\Theta}$}$
%\begin{mathpar}
%\inferrule[k-opcon]{
%	\tKind{\tvarCtx}{\fvalCtx}{\taudef}{\karrow{\kappaidx}{\karrow{\klist{\kDen}}{(\ksum{\kDen}{\kunit})}}}
%}{
%	\opType{\tvarCtx}{\fvalCtx}{{\tops{op}{\kappaidx}{i}{a}{\taudef}}}{
%	\kOpS{op}{\kappaidx}}
%}
%
%\inferrule[k-opcons]{
%	\opType{\tvarCtx}{\fvalCtx}{\theta_1}{\Theta_1}\\
%	\opType{\tvarCtx}{\fvalCtx}{\theta_2}{\Theta_2}\\\\
%	\text{dom}(\theta_1) \cap \text{dom}(\theta_2) = \emptyset
%}{
%	\opType{\tvarCtx}{\fvalCtx}{\theta_1; \theta_2}{\Theta_1, \Theta_2}
%}
%\end{mathpar}
%$\fbox{$\exprOKX{e}$}$
%\begin{mathpar}
%\inferrule[k-e-var]{ }{
%	\exprOK{\tvarCtx}{\fvalCtx}{\evar{x}}
%}
%
%%\inferrule[k-e-lam]{
%%	\tKindX{\tau}{\kTypeBlur}\\
%%	\exprOK{\tvarCtx}{\fvalCtx}{e}
%%}{
%%	\exprOKX{\elam{\evar{x}}{\tau}{e}}
%%}
%
%\inferrule[k-e-op]{
%	\fval{Tycon}{-}{-}{\theta} \in \fvalCtx\\
%	\tops{op}{\kappaidx}{i}{a}{-} \in \theta\\
%	\tKindX{\tauidx}{\kappaidx}\\
%	\exprOKX{e_1}\\
%	\cdots\\
%	\exprOKX{e_n}
%}{
%	\exprOKX{\eop{Tycon}{op}{\tauidx}{\splat{e}{1}{n}}}
%}
%\end{mathpar}
%$\fbox{$\kEq{\kappa}$}$
%\begin{mathpar}
%\inferrule[t-eq]{ }{
%	\kEq{\kTypeBlur}
%}
%
%\inferrule[i-eq]{ }{
%	\kEq{\kint}
%}
%
%\inferrule[l-eq]{ }{
%	\kEq{\klabel}
%}
%
%\inferrule[list-eq]{
%	\kEq{\kappa}
%}{
%	\kEq{\klist{\kappa}}
%}
%\\
%\inferrule[u-eq]{ }{
%	\kEq{\kunit}
%}
%
%\inferrule[p-eq]{
%	\kEq{\kappa_1}\\
%	\kEq{\kappa_2}
%}{
%	\kEq{\kpair{\kappa_1}{\kappa_2}}
%}
%
%\inferrule[s-eq]{
%	\kEq{\kappa_1}\\
%	\kEq{\kappa_2}
%}{
%	\kEq{\ksum{\kappa_1}{\kappa_2}}
%}
%\end{mathpar}
%\caption{\small Kinding for type and operator constructors and external terms, and equality kinds (see text).  Type-level terms stored in the constructor context are not needed during kinding, indicated using a dash.}
%\label{kindprog}
%\vspace{-10pt}
%\end{figure}
%\begin{figure}[t]
%\small
%%\vspace{-15pt}
%$\fbox{\inferrule{}{\tKindX{\tau}{\kappa}}}$
%%~~~$\chi ::= \cdot \pipe \checkmark$
%\begin{mathpar}
%\small\inferrule[k-var]{
%}{
%  \tKind{\tvarCtxX{t}{\kappa}}{\fvalCtx}{\tvar{t}}{\kappa}
%}
%~~~~
%\inferrule[k-arrow-i]{
%  \tKind{\tvarCtxX{t}{\kappa_1}}{\fvalCtx}{\tau}{\kappa_2}
%}{
%  \tKindX{\tlam{t}{\kappa_1}{\tau}}{\karrow{\kappa_1}{\kappa_2}}
%}
%~~~~
%\inferrule[k-arrow-e]{
%  \tKindX{\tau_1}{\karrow{\kappa_1}{\kappa_2}}\\\\
%  \tKindX{\tau_2}{\kappa_1}
%}{
%  \tKindX{\tapp{\tau_1}{\tau_2}}{\kappa_2}
%}
%
%\text{\color{gray} (kinding for integers, labels, lists, products and sums also standard)}
%
%%%
%%%\inferrule{ }{
%%%	\tKindX{\tstr{str}}{\kstr}
%%%}(\text{str}^I_\tau)
%%%
%%%\inferrule{ }{
%%%	\tKindX{\tunit}{\kunit}
%%%}(\text{1}^I_\tau)
%%%
%%%\inferrule{
%%%	\tKindX{\tau_1}{\kappa_1}\\
%%%	\tKindX{\tau_2}{\kappa_2}
%%%}{
%%%	\tKindX{\tpair{\tau_1}{\tau_2}}{\kpair{\kappa_1}{\kappa_2}}
%%%}({\times}^I_\tau)
%%%
%%%\inferrule{
%%%	\tKindX{\tau}{\kpair}{\kappa_1}{\kappa_2}
%%%}{
%%%	\tKindX{\tfst{\tau}}{\kappa_1}
%%%}({\times}^{E1}_\tau)
%%%
%%%\inferrule{
%%%	\tKindX{\tau}{\kpair}{\kappa_1}{\kappa_2}
%%%}{
%%%	\tKindX{\tsnd{\tau}}{\kappa_2}
%%%}({\times}^{E2}_\tau)
%%%\inferrule{ }{
%%%\inferrule{
%%%	\tKindX{\tau_{1}}{\kappa_{1}}\\
%%%	\tKindX{\tau_{2}}{\kappa_{2}}
%%%}{
%%%	\tKindX{\tpair{\tau_{1}}{\tau_{2}}}{\kpair{\kappa_{1}}{\kappa_{2}}}
%%%}~(\times_\tau)
%%%
%%%\inferrule{
%%%	\tKindX{\tau}{\kpair{\kappa_{1}}{\kappa_{2}}}
%%%}{
%%%	\tKindX{\tfst{\tau}}{\kappa_{1}}
%%%}~(\text{fst}_\tau)
%%%
%%%\inferrule{
%%%	\tKindX{\tau}{\kpair{\kappa_{1}}{\kappa_{2}}}
%%%}{
%%%	\tKindX{\tsnd{\tau}}{\kappa_{2}}
%%%}~(\text{snd}_\tau)
%%%
%%% TODO: list
%\inferrule[k-eq]{
%	\kEq{\kappa}\\
%	\tKindX{\tau_1}{\kappa}\\
%	\tKindX{\tau_2}{\kappa}\\\\
%	\tKindX{\tau_3}{\kappa'}\\
%	\tKindX{\tau_4}{\kappa'}
%}{
%	\tKindX{\tifeq{\tau_1}{\tau_2}{\kappa}{\tau_3}{\tau_4}}{\kappa'}
%}
%
%\inferrule[k-ty-i]{
%	\fval{Tycon}{\kappaidx}{-}{-} \in \fvalCtx\\\\
%	\tKind{\tvarCtx}{\fvalCtx}{\tauidx}{\kappaidx}
%}{
%	\tKindX{\ttype{Tycon}{\tauidx}}{\kTypeBlur}
%}
%
%\inferrule[k-ty-e]{
%	\tKindX{\tau}{\kTypeBlur}\\
%	\fval{Tycon}{\kappaidx}{-}{-} \in \fvalCtx\\\\
%	\tKind{\tvarCtxX{x}{\kappaidx}}{\fvalCtx}{\tau_1}{\kappa}\\
%	\tKindX{\tau_2}{\kappa}
%}{
%	\tKind{\tvarCtx}{\fvalCtx}{\tfamcase{\tau}{Tycon}{x}{\tau_1}{\tau_2}}{\kappa}
%}
%~~~~~~
%\inferrule[k-d-i]{
%	\tKindX{\tau}{\kTypeBlur}\\\\
%	\isItermX{\ibar}
%}{
%	\tKindX{\tden{\ibar}{\tau}}{\kDen}
%}
%
%\inferrule[k-d-i-checked]{
%	\tKind{\Delta}{\fvalCtx}{\tau}{\kTypeBlur}\\\\
%	\isItermX{\ibar}
%}{
%	\tKind{\Delta}{\fvalCtx}{\tden{\ibar}{\tau}^\checkmark}{\kDen}
%}
%~~~~~~~
%\inferrule[k-d-e]{
%	\tKindX{\tau}{\kDen}
%}{
%	\tKindX{\ttypeof{\tau}}{\kTypeBlur}
%}
%%
%%\inferrule[iterm-intro]{
%%	\isIterm{\tvarCtx}{\emptyctx}{\fvalCtx}{\iota}
%%}{
%%	\tKindX{\titerm{\iota}}{\kITerm}
%%}
%
%\inferrule[k-reptype-i]{
%	\isItype{\tvarCtx}{\fvalCtx}{\sbar}
%}{
%	\tKindX{\titype{\sbar}}{\kIType}
%}
%\end{mathpar}
%$\fbox{$\isItermX{\ibar}$}$
%\begin{mathpar}
%\inferrule[k-i-var]{ }{
%	\isItermX{\evar{x}}
%}
%~~~~~~
%\inferrule[k-i-fix]{
%	\isItypeX{\sbar}\\
%	\isItermX{\ibar}
%}{
%	\isItermX{\ifix{\evar{x}}{\sbar}{\ibar}}
%}
%~~~~~~
%\inferrule[k-i-transof]{
%	\tKindX{\tau}{\kDen}
%}{
%	\isItermX{\itransof{\tau}}
%}
%%\inferrule[i-lam-kinding]{
%%	\isItypeX{\sbar}\\\\
%%	\isItermX{\ibar}
%%}{
%%	\isItermX{\ilam{\evar{x}}{\sbar}{\ibar}}
%%}
%
%\text{\color{gray} (omitted rules are trivially recursive, like \textsc{k-i-fix})}
%%\inferrule[iterm-dereify-kinding]{
%%	\tKindX{\tau}{\kITerm}
%%}{
%%	\isItermX{\iup{\tau}}
%%}
%%
%\end{mathpar}
%$\fbox{$\isItypeX{\sbar}$}$
%\begin{mathpar}
%%\inferrule[i-int-kinding]{ }{
%%	\isItypeX{\dint}
%%}
%%
%%\inferrule[i-prod-kinding]{
%%	\isItypeX{\sigma_1}\\\\
%%	\isItypeX{\sigma_2}
%%}{
%%	\isItypeX{\dpair{\sigma_1}{\sigma_2}}
%%}
%%
%\inferrule[k-s-unquote]{
%	\tKindX{\tau}{\kIType}
%}{
%	\isItypeX{\dup{\tau}}
%}
%
%\inferrule[k-s-repof]{
%	\tKindX{\tau}{\kTypeBlur}
%}{
%	\isItypeX{\trepof{\tau}}
%}
%
%\text{\color{gray} (omitted forms are trivially recursive)}
%\end{mathpar}
%\caption{\small Kinding for type-level terms. The kinding context $\tvarCtx$ maps type variables to kinds. Checked derivates (introduced by rule \textsc{k-d-i-checked}, discussed in Sec. \ref{repcon}) should only be constructed by the semantics (this can be enforced by a parser, so we will assume it in the metatheory).}
%\label{tlkind}
%\vspace{-10px}
%\end{figure} 
%In most languages, types are formed by applying one of a collection of \emph{type constructors} to zero or more \emph{indices}. In @$\lambda$, the situation is notionally similar. User-defined type constructors can be declared at the top of a program (or lifted to the top, in practice) using \textsf{tycon}. Each constructor in the program must have a unique name, written e.g. \fvar{Nat}.\footnote{We assume naming conflicts can be avoided by some extrinsic mechanism.} A type constructor must also declare an \emph{index kind}, $\kappaidx$. A type is introduced by applying a type constructor to an index of this kind, written $\ttype{Tycon}{\tauidx}$. 
%
%@$\lambda$ supports, and makes extensive use of, simply-kinded type-level computation. Specifically, type-level terms, $\tau$, themselves form a typed lambda calculus. The classifiers of type-level terms are called \emph{kinds}, $\kappa$, to distinguish them from  \emph{types}. The kinding rules are specified in Fig. \ref{kindprog} and Fig. \ref{tlkind}. Types are  type-level values of kind $\kTypeBlur$, introduced as just described. The kind $\kTypeBlur$ also has an elimination form, $\tfamcase{\tau}{Tycon}{x}{\tau_1}{\tau_2}$, allowing the extraction of a type index by case analysis against a contextually-available type constructor. To a first approximation, one might think of type constructors as constructors of a built-in open datatype, $\kTypeBlur$, at the type-level. Like open datatypes, there is no notion of exhaustiveness so the default case is required for totality.
%%We will write the kind of types as $\kTypeBlur$, though it is also written $\star$ or \verb|Type| in various similarly structured languages (see Sec. \ref{related-work}). 
%% Rather than there being a fixed set of type constructors, we allow the programmer to declare new type  constructors, and give the static and dynamic semantics of their associated operators, by writing type-level functions. In the semantics for this calculus, our kind system combined with techniques borrowed from the typed compilation literature and a form of type abstraction allow us to prove strong type safety, decidability and conservativity theorems.
%
%
% To permit the implementation of interesting type systems, the type-level language includes several kinds other than $\kTypeBlur$. We lift several  standard functional fragments to the type level: unit ($\kunit$), binary products ($\kpair{\kappa_1}{\kappa_2}$), binary sums ($\ksum{\kappa_1}{\kappa_2}$), lists ($\klist{\kappa}$) and integers ($\dint$). We also include labels ($\klabel$), written in a slanted font, e.g. $\tlabel{myLabel}$, which are string-like values that only support comparison and play a distinguished role in the expanded syntax, as we will later discuss. Our first example, $\fvar{Nat}$, is indexed trivially, i.e. by unit kind, $\kunit$, so there is only one natural number type, $\ttype{Nat}{\tunit}$, but we will show examples of type constructors that are indexed in more interesting ways in later portions of this work. For example, $\fvar{LabeledTuple}$ has index kind $\klist{\kpair{\klabel}{\kTypeBlur}}$. The type constructor $\fvar{Arrow}$ is included in the initial constructor context, $\fvalCtx_0$, shown in Fig. \ref{arrow}, and has index kind $\kpair{\kTypeBlur}{\kTypeBlur}$.
% 
% Type constructors are not first-class; they do not themselves have arrow kind as in some kind systems (e.g.  \cite{watkins2008specifying}; Ch. 22 of \emph{PFPL} describes a related system \cite{pfpl}). The type-level language does, however, include total functions of conventional arrow kind, written $\karrow{\kappa_1}{\kappa_2}$. Type constructor application can be wrapped in a type-level function to emulate a first-class or uncurried version of a type constructor for convenience (indeed, such a wrapper could be generated automatically, though we do not do so). 
%
%Two type-level terms of kind $\kTypeBlur$ are equivalent if they apply the same constructor, identified by name, to equivalent indices. Going further, we ensure that deciding type equivalence requires only checking for syntactic equality after normalization by imposing the restriction that equivalence at a type constructor's index kind must be decidable in this way. Our treatment of equivalence in the type-level language is thus quite similar to the treatment of term-level equality using ``equality types'' in a language like Standard ML. A kind $\kappa$ is an  \emph{equality kind} if $\kEq{\kappa}$ can be derived (Fig. \ref{kindprog}). Conditional branching on the basis of equality at an equality kind can be performed in the type-level language (e.g. in \tvar{ifeq}, Fig. \ref{helpers}, discussed below). Equivalence at arrow kind is not decidable by our criteria, so type-level functions cannot appear within type indices. This also prevents general recursion from arising at the type level. Without this restriction, a type-level function taking a type as an argument could ``smuggle in'' a self reference as a type index, extracting it via case analysis (continuing our analogy to open datatypes, this is closely related to the positivity condition for inductive datatypes in total functional languages like Coq).% as maintaining the metatheoretic guarantee that typing respects type equivalence would impose a substantial burden in such a setting.% (a na\"ive approach to this would impose non-trivial extrinsic proof obligations onto extension developers that, unlike in others in this thesis, could threaten type safety).
%
%Every type constructor also defines a \emph{schema}, a type-level function that associates with every type an internal \emph{representation type}. We will return to this in Sec. \ref{repcon} after introducing operators. 
%\subsection{Operators and Operator Constructors}\label{opcons}
%\begin{figure}[t]
%\small
%\[
%\begin{array}{rcl}
%\tvar{failure} & := & \tinr{\kDen}{\tunit}\\
%\tvar{ok} & := & (\tlam{d}{\kDen}{\tinl{\kunit}{\tvar{d}}})\\
%\tvar{ifeq} & := & (\tlam{t1}{\kTypeBlur}{\tlam{t2}{\kTypeBlur}{\tlam{d}{\kDen}{
%	\\& & \quad \tifeq{\tvar{t1}}{\tvar{t2}}{\kTypeBlur}{\tvar{ok}~\tvar{d}}{\tvar{failure}}\\
%\tvar{arity0} & := & (\tlam{a}{\klist{\kDen}}{\tlam{d}{(\ksum{\kDen}{\kunit})}{\tfold{\tvar{a}}{\tvar{d}}{\_}{\_}{\_}{\tvar{failure}}}})\\
%\tvar{arity1} & := & (\tlam{a}{\klist{\kDen}}{\tlam{k}{\karrow{\kDen}{(\ksum{\kDen}{\kunit})}}{
%	\\ & & \quad \tfold{\tvar{a}}{\tvar{tyerr}}{h}{t}{\_}{
% \tvar{arity0}~\tvar{t}~(\tvar{k}~\tvar{h})}}})\\
%\tvar{arity2} & := & (\tlam{a}{\klist{\kDen}}{\tlam{k}{\karrow{\kDen}{\karrow{\kDen}{(\ksum{\kDen}{\kunit})}}}{
%	\\ & & \quad \tfold{\tvar{a}}{\tvar{tyerr}}{h}{t}{\_}{
% \tvar{arity1}~\tvar{t}~(\tvar{k}~\tvar{h})}}})\\
% \tvar{arity3} & := & (\tlam{a}{\klist{\kDen}}{\tlam{k}{\karrow{\kDen}{\karrow{\kDen}{\karrow{\kDen}{(\ksum{\kDen}{\kunit})}}}}{
%	\\ & & \quad \tfold{\tvar{a}}{\tvar{tyerr}}{h}{t}{\_}{
% \tvar{arity2}~\tvar{t}~(\tvar{k}~\tvar{h})}}})
%}}}
%\end{array}
%\]
%\caption{\small Useful (type-level) functions for writing operator definitions.}
%\vspace{-10pt}
%\label{helpers}
%\end{figure}
%\begin{figure}[t]
%\small
%\[
%\begin{array}{rcl}
%\Phi_0 & := & \emptyset, \fval{Arrow}{\kpair{\kTypeBlur}{\kTypeBlur}}{\tau_0}{\theta_0}\\
%\tau_0 & := & \tlam{i}{\kpair{\kTypeBlur}{\kTypeBlur}}{\titype{\darrow{\trepof{\tfst{\tvar{i}}}}{\trepof{\tsnd{\tvar{i}}}}}}\\
%\theta_0 & := & \tops{ap}{\kunit}{i}{a}{
%	\tlam{i}{\kunit}{\tlam{a}{\klist{\kDen}}{
%		\tvar{arity2}~\tvar{a}~\tlam{d1}{\kDen}{\tlam{d2}{\kDen}{
%		\\& & \quad \tfamcase{\ttypeof{\tvar{d1}}}{Arrow}{x}{
%		\\& & \quad\quad \tvar{ifeq}~\tfst{\tvar{x}}~\ttypeof{\tvar{d2}}
%		\\& & \quad\quad\quad \tden{\itransof{\tvar{d1}}~\itransof{\tvar{d2}}}{\tsnd{\tvar{x}}}
%		\\& & \quad\hspace{-3px}}{\tvar{tyerr}}
%		}}
%	}}
%}
%\end{array}
%\]
%\caption{\small The initial constructor context, $\fvalCtx_0$, defines the $\fvar{Arrow}$ type constructor. The introductory form, $\elam{x}{\tau}{e}$, is built into the language but the elimination form goes through the operator constructor $\fvar{Arrow}.\opvar{ap}$.}
%\label{arrow}
%\end{figure}
%User-defined operator constructors are declared using \textsf{opcon}.  For reasons that we will discuss, our calculus associates every operator  constructor with a type constructor. The \emph{fully-qualified name} of every operator constructor, e.g. $\fvar{Nat}.\opvar{z}$, must be unique. Operator constructors, like type constructors, declare an index kind, $\kappaidx$. In our first example, all the operator constructors are indexed trivially (by index kind $\kunit$), but other examples use more interesting indices. For example, $\fvar{LabeledTuple}.\opvar{prj}$ is the operator constructor used to access a field of a labeled tuple, so it has index kind $\klabel$. An operator itself is, notionally, selected by indexing an operator constructor, e.g. $\fvar{Nat}.\opvar{s}\langle \tunit \rangle$, but technically neither operator constructors nor operators are first-class at any level (additional machinery would be needed, e.g. an \textsf{Op} kind, but this is not fundamental to our calculus). 
%Instead, in the external language, an operator constructor is applied by simultaneously providing an index and  $n \geq 0$ \emph{arguments}, written $\eop{Tycon}{op}{\tauidx}{\splat{e}{1}{n}}$\footnote{It may be helpful to distinguish between type/operator constructors and \emph{term formers}, which are the metatheoretic entities that ultimately define the syntax. There are term formers at all levels in the calculus. For example, operator constructor application and $\lambda$ are  external term formers, and type constructor application is a type-level term former. An abstract syntax following Harper's conventions highlights this distinction \cite{pfpl}: the term formers $\mathtt{lam}$, $\mathtt{tcapp}$ and $\mathtt{ocapp}$, define, respectively the forms $\mathtt{lam}[\tau](x.e)$, $\texttt{tcapp}[\fvar{Tycon}](\tau)$ and $\mathtt{ocapp}[\fvar{Tycon}, \opvar{op}, \tauidx](\splat{e}{1}{n})$.}. For example, on line 18 of Fig. \ref{nat}, we see the operator constructors $\fvar{Nat}.\opvar{z}$ and $\fvar{Nat}.\opvar{s}$ being applied to compute $two$. In SML, the projection operator \verb|#3| can be applied to an $n$-tuple, $e$, iff $n \geq 3$. Note that it thus cannot be a function with a standard arrow type. Notionally, \verb|#| is an operator constructor and \verb|3| is its index. In an active embedding of $n$-tuples into @$\lambda$, this would be written $\eop{Tuple}{prj}{3}{e}$ (we will nearly recover ML's syntax later). %\footnote{Although our focus here is entirely on semantics, a brief note on syntax: in the expanded syntax, the trivial indices and empty argument lists can be omitted, so we could write \texttt{Nat.s(Nat.s(Nat.z))}. With the ability to ``open'' a type's operators into the context, we could shorten this still to \texttt{s(s(z))}. Alternatively, with the ability to define a TSL in a manner similar to that in Sec. \ref{aparsing}, we might instead just write \texttt{2}.}
%
%\subsection{Operator Definitions and Derivates}
%The expressive and metatheoretic power of the calculus arises from how the rules for the active typing judgement handle this  operator constructor application form. Rather than fixing the specification of a finite collection of operator constructors and tasking the \emph{compiler} with deciding a typing derivation on its basis, the specification instead tasks the \emph{operator constructor's definition}: {a type-level function that must decide the type assignment and the translation on the basis of the index and arguments, or decide that this is not possible}. As indicated by rule \textsc{k-opcon} in Fig. \ref{kindprog}, an operator constructor with index kind $\kappaidx$ must have a definition of kind $$\kappaidx \rightarrow \klist{\kDen} \rightarrow (\kDen + \kunit)$$ As this suggests, the active typing judgement  will provide the operator index and a list of {recursively determined \emph{derivates}}, which have kind $\kDen$, for the arguments and ask the definition to return a value of ``option kind'', $\ksum{\kDen}{\kunit}$, where the trivial case indicates that a derivate cannot be constructed due to an invalid index, an incorrect number of arguments or an argument of an invalid type.\footnote{In practice, we would require operator constructor providers to report information about the precise location of the error (e.g. which argument was of incorrect type) and provide an appropriate error message and other metadata, but we omit this in the semantics.}
%
%A \emph{derivate} is introduced in operator definitions by the form $\tden{\ibar}{\tau}$, where $\tau$ is a type assignment and has kind $\kTypeBlur$ and $\ibar$ is a \emph{translational internal term}. The syntax for translational internal terms mirrors that of internal terms, $\iota$, adding an additional form, $\itransof{\tauD}$, that permits one translation to be composed using the translations of other derivates (e.g. the derivate of an argument). Fig. \ref{tlkind} shows the kinding rules for derivates and translational internal terms (we will discuss checked derivates shortly; they are only used internally by the semantics). The type assignment can be extracted from a derivate using $\ttypeof{\tauD}$.
%
%Let us return to Fig. \ref{nat} to build intuition. 
%The definition of $\fvar{Nat}.\opvar{z}$ in Fig. \ref{nat} is quite simple: it returns the derivate $\tden{{0}}{\ttype{Nat}{\tunit}}$ if no arguments were provided, and indicates an error otherwise. This is done by calling a simple helper function, $\tvar{arity0}$, seen in Fig. \ref{helpers}. The definition of $\fvar{Nat}.\opvar{s}$ is only slightly more complex, because it requires inspecting a single argument. The helper function $\tvar{arity1}$ checks the arity of the argument list. If it is correct, it passes the single derivate to the shown continuation, returning the error case otherwise (which we give a more descriptive name, $\tvar{failure}$). The continuation we provided checks that the type of the argument is $\ttype{Nat}{\tunit}$ using a similar helper function, $\tvar{ifeq}$. If so, it constructs the derivate $\tden{\itransof{\tvar{d}}+1}{\ttype{Nat}{\tunit}}$ and $\tvar{ifeq}$ wraps it with $\tvar{ok}$. We will return to the recursor shortly.
%
%These two operator definitions make the following rules admissible in a constructor context, let us call it $\fvalCtx_{\mathbf{T}}$, containing the initial constructor context (Fig. \ref{arrow}) and the constructors of  Fig. \ref{nat} and any typing context well-kinded under that constructor context (the rules governing context kinding are in the appendix):
%\newcommand{\fvalT}{\fvalCtx_{\mathbf{T}}}
%\begin{mathpar}
%\inferrule[gt-z]{ }{
%	\ecompiles{\Upsilon}{\fvalT}{\eop{Nat}{z}{\tunit}{}}{\ttype{Nat}{\tunit}}{0}
%}
%
%\inferrule[gt-s]{
%	\ecompiles{\Upsilon}{\fvalT}{e}{\ttype{Nat}{\tunit}}{\iota}
%}{
%	\ecompiles{\Upsilon}{\fvalT}{\eop{Nat}{s}{\tunit}{e}}{\ttype{Nat}{\tunit}}{\iota + 1}
%}
%\end{mathpar}
%
% We will consider the complexities shortly, but it requires little more than basic intuition to see that we can strongly embed the types and introductory forms of G\"odel's $\mathbf{T}$ in @$\lambda$ quite directly (adding essentially only a barrage of trivial indices which will be removed syntactically in the expanded syntax). If we keep the constructor context fixed, we should be able to prove the embedding correct inductively. We will define this embedding later\todo{refer to it now}.
%
% Figure \ref{att} shows how $\fvalCtx_{\mathbf{T}}$ would be built up and how these conclusions could be derived. The active typing judgement simply defers to the \emph{checked active typing judgement} (Fig. \ref{absatt}) to first produce a type assignment and a translational internal term, called the \emph{abstract translation}. It then \emph{deabstracts} the abstract translation to produce a type assignment and translation (Fig. \ref{deabstraction}). 
% 
% The rule \textsc{att-op} shows how operator constructor definitions are invoked. First, type assignments  and translational internal terms are recursively derived. Then, the operator constructor definition, $\taudef$ is extracted from the constructor context and applied to the index, $\tauidx$, and a list of \emph{checked derivates} constructed from the derivations produced in the first step. If this produces a derivate, $\tauD$, via the left case of the sum, then it is itself, finally, \emph{checked} to produce a type assignment, $\tau$, and translational internal term, $\ibar$. 
% 
%The normalization rules for type-level terms are shown in Fig. \ref{tleval} and are rather uninteresting. We will discuss their metatheoretic properties in Sec. \ref{theory}.
% 
% Derivate checking, shown in Fig. \ref{ait}, is described in the next two subsections. 
% Note that derivate checks are still static checks; they do not change the translation. Derivate checking is sufficient to guarantee type safety and preclude interfere issues between extensions, as we will describe, because taking a path through an operator constructor definition that produces a problematic derivate will fail. If providers do not want to burden clients with this variety of error, which would require them to debug the constructor definition itself, we will discuss how constructor correctness could be verified mechanically by strengthening the kind system in Sec. \ref{limitations}. %We will discuss how to prove that an extension  operator definition could be \emph{proven} to not produce such derivates (so that users only ever face type errors arising from their own code, not also extension correctness issues) in Sec. \ref{limitations}. The checks are sophisticated enough that they cannot be eliminated by a simple kind system.  
%
%%$\tvar{pop\_final} : \karrow{\karrow{\klist{\kDen}}{(\karrow{\kTypeBlur}{\karrow{\kITerm}{(\ksum{\kDen}{\kunit})})}}}{(\ksum{\kDen}{\kunit})}$ ``pops'' a derivate from the head of the list and, if no other arguments remain, passes its translation and type to the ``continuation'', returning the error case otherwise. The continuation checks if the argument type is equal to $\ttype{Nat}{\tunit}$ using $\tvar{check\_type} : \karrow{\kTypeBlur}{\karrow{\kTypeBlur}{\karrow{\kDen}{(\ksum{\kDen}{\kunit})}}}$, which operates similarly. If these arity and type checks succeed, the resulting derivate is composed by adding one to the translation of the argument, passed into the continuation as $\tvar{x}$ in our example, and pairing it with the natural number type: $\tden{\titerm{\iup{\tvar{x}} + 1}}{\ttype{Nat}{\tunit}}$. We will return to the definition of $\fvar{Nat}.\opvar{rec}$ after in the next subsection.
%
%%We leave adding the ability to declare and manipulate new contexts as future work.
% 
%\subsection{Representational Consistency Implies Type Safety}\label{repcon}
%\begin{figure}[t]
%\small
%$\fbox{\inferrule{}{\pcompilesX{\rho}}}$
%\begin{mathpar}
%\inferrule[att-tycon]{
%	\pcompiles{\fvalCtxX{\fvalDf}}{\rho}{\iota}
%}{
%	\pcompilesX{\pfam{\familyDf}{\rho}}
%}
%%
%%\inferrule[att-def]{
%%	\tEvalX{\tau}{\tau'}\\
%%	\pcompilesX{\subst{\tau'}{\tvar{t}}{\rho}}
%%}{
%%	\pcompilesX{\pdef{t}{\kappa}{\tau}{\rho}}
%%}
%
%\inferrule[att-e-prog]{
%	\ecompiles{\emptyctx}{\fvalCtx}{e}{\tau}{\iota}%\\
%%	\eraseX{\gabs}{\iota}
%}{
%	\pcompilesX{e}
%}
%\end{mathpar}
%$\fbox{\inferrule{}{\ecompilesX{e}{\tau}{\iota}}}$
%~~~$\etCtx ::= \emptyctx \pipe \etCtxX{x}{\tau}$
%\begin{mathpar}
%\inferrule[att-e]{
%	\ecompilesA{\Upsilon}{\fvalCtx}{e}{\tau}{\ibar}\\
%	\eraseX{\ibar}{\iota}
%}{
%	\ecompiles{\Upsilon}{\fvalCtx}{e}{\tau}{\iota}\\
%}
%\end{mathpar}
%\caption{\small Active typing judgement for programs and external terms. The deabstraction ($\lightning$) judgement is in Fig. \ref{deabstraction}.}
%\label{att}
%%\vspace{-10pt}
%\end{figure}
%\begin{figure}[t]
%\small
%$\fbox{\inferrule{}{\ecompilesAX{e}{\tau}{\ibar}}}$
%\begin{mathpar}
%\inferrule[att-var]{ }{
%	\ecompilesA{\etCtxX{x}{\tau}}{\fvalCtx}{\evar{x}}{\tau}{\evar{x}}
%}
%
%\inferrule[att-lam]{
%	\tEvalX{\tau_1}{\tau_1'}\\
%%	\ddbar{\fvar{Arrow}}{\fvalCtx}{\trepof{\tau_1'}}{\sbar_1}\\
%	%\delfromtau{$\Xi_0$}{\fvalCtx}{\tau_1'}{\sabs}\\\\
%	\ecompilesA{\etCtxX{x}{\tau_1'}}{\fvalCtx}{e}{\tau_2}{\ibar}
%}{
%	\ecompilesA{\etCtx}{\fvalCtx}{\elam{\evar{x}}{\tau_1}{e}}{\ttype{Arrow}{\tpair{\tau_1'}{\tau_2}}}{\ilam{\evar{x}}{\trepof{\tau_1'}}{\ibar}}
%}
%
%\inferrule[att-op]{
%%	\tEvalX{\tauidx}{\tau_{\text{idx}}'}\\\\
%	\ecompilesA{\etCtx}{\fvalCtx}{e_1}{\tau_1}{\ibar_1}~~~~
%	\cdots~~~~
%	\ecompilesA{\etCtx}{\fvalCtx}{e_n}{\tau_n}{\ibar_n}\\
%	\fval{Tycon}{-}{-}{\theta} \in \fvalCtx\\
%	\tops{op}{\kappaidx}{i}{a}{\taudef} \in \theta\\
%	\tEvalX{\taudef~\tauidx~(\tden{\ibar_1}{\tau_1}^\checkmark :: \cdots :: \tden{\ibar_n}{\tau_n}^\checkmark :: \tnil{\kDen})}{\tinl{\kunit}{\tauD}}\\
%	\dcheck{\Upsilon}{\fvalCtx}{\fvar{Arrow}, \fvar{Tycon}}{\tauD}{\tau}{\ibar}
%}{
%	\ecompilesA{\etCtx}{\fvalCtx}{\eop{Tycon}{op}{\tauidx}{\splat{e}{1}{n}}}{\tau}{\ibar}
%}
%\end{mathpar}
%\caption{\small The checked active typing judgement. The normalization judgement for type-level terms ($\Downarrow$) is specified in Fig. \ref{tleval}. Derivate checking is specified in Fig. \ref{ait}. Note that checked derivates can \emph{only} be constructed by the \textsc{att-op} rule (we enforce this judgementally in Sec. \ref{theory}).}
%\label{absatt}
%\end{figure}
%In our example, $\fvar{Nat}.\opvar{z}$ produces a translation of internal type $\dint$. For the translation produced by $\fvar{Nat}.\opvar{s}$ to be well-typed requires that $\itransof{\tvar{d1}}$ be of internal type $\dint$. Because we have checked that $\ttypeof{\tvar{d1}}$ is $\ttype{Nat}{\tunit}$ and the only other introductory form is $\fvar{Nat}.\opvar{z}$, this is not an issue. Well-typed external terms of type $\ttype{Nat}{\tunit}$ will always translate to well-typed internal terms of internal type $\dint$ as given.
%
%This would not be true if, for example, we added an operator constructor $\fvar{Nat}.\opvar{z2}$ that produced the derivate $$\tden{{(0, ())}}{\ttype{Nat}{\tunit}}$$
%\noindent
%In this case, there would be two different internal types, $\dint$ and $\dpair{\dint}{\dunit}$,  associated with a single external type, $\ttype{Nat}{\tunit}$. This makes it impossible to reason \emph{compositionally} about the translation of an external term of type $\ttype{Nat}{\tunit}$, so our implementation of $\fvar{Nat}.\opvar{s}$ would produce ill-typed translations in some cases but not others. Similarly, we wouldn't be able to write functions over all natural numbers because there would not be a well-typed translation to give to such a function. This violates type safety: there are now well-typed external terms, according to the active typing judgement, for which evaluating the corresponding translation would ``go wrong''. 
%
%To reason compositionally about the semantics of well-typed external terms when they are given meaning by translation to a typed internal language, we must have that the statics maintains the following property: for every  type, $\tau$, there must exist an internal type, $\sigma$, called its \emph{representation type}, such that the translation of every external term of type $\tau$ has internal type $\sigma$. This principle of \emph{representational consistency} arises essentially as a strengthening of the inductive hypothesis necessary to prove that all well-typed external terms translate to well-typed internal terms, precisely because $\lambda$ and operators like $\fvar{Nat}.\opvar{s}$ are defined compositionally. It is closely related to the concept of \emph{type-preserving compilation} developed by Morrisett et al. for the TIL compiler for SML \cite{tarditi+:til-OLD}. %That is, instead of being used as a necessary condition for compiler correctness, we are using it as a sufficient condition for type safety. 
%%We emphasize the distinction between translation (which in our calculus endows terms with a dynamic semantics) and compilation (which must preserve the dynamic semantics of terms). %In the presence of extensions, checking for representational consistency becomes subtle if we wish to guarantee conservativity, as we will discuss in the next subsection.
%
%For the semantics to ensure that representational consistency is maintained by all operator definitions, we require, as briefly mentioned in Sec. \ref{types}, that each type constructor must declare a \emph{representation schema} after the keyword \textsf{schema}. This must be a type-level function of kind $\karrow{\kappaidx}{\kIType}$, where $\kappaidx$ is the index kind of the type constructor. When the semantics (i.e. the compiler) needs to determine the representation type of the type $\ttype{Tycon}{\tauidx}$ it simply applies the representation schema of $\fvar{Tycon}$ to ${\tauidx}$.
%
%The kind $\kIType$ has introductory form $\titype{\bar \sigma}$ and no elimination form. It is not an equality kind. There are two forms of \emph{translational internal type}, $\bar \sigma$, that do not correspond to forms in $\sigma$ that, as with translational internal terms, allow  compositional formation:
%\begin{enumerate}
%\item $\dup{\tau}$ ``unquotes'' the translational internal type $\tau$, so that internal types can be formed programmatically\footnote{An analagous form could be added to translational internal terms, but the same effect can be achieved by working directly with derivates, albeit with some inconvenience when ``temporary intermediates'' are needed.} \todo{example}
%\item $\trepof{\tau}$ refers to the representation type of type $\tau$. This type may have come from an argument derivate or extracted from a type or operator index.
%\end{enumerate}
%
%These additional forms are not needed by the representation schema of $\fvar{Nat}$  because it is trivially indexed. In Fig. \ref{tuple}, we show an example of the type constructor $\fvar{Tuple}$, implementing the semantics of $n$-tuples by translation to nested binary products. Here, the representation schema requires referring to the representations of the tuple's constituent types, given in the type index\todo{add this example from appendix of ESOP}.
%
%Translational internal terms also contain translational internal types. We see this in the definition of the recursor on natural numbers, $\fvar{Nat}.\opvar{rec}$. After checking the arity, extracting the derivates of its three arguments, and checking their types (using \fvar{Arrow} to bind variables in the third position), it produces a derivate that implements the necessarily recursive dynamic semantics using a fixpoint computation in the internal language. This derivate's type assignment is the arbitrary type, $\tvar{t2}$. We cannot know what the representation type of $\tvar{t2}$ is (even in a closed constructor context, because the schema is not directly callable), so we refer to it abstractly using $\trepof{\tvar{t2}}$. Luckily, the translation is well-typed \emph{independent of the representation type}. In other words, a proof of representational consistency of the derivate produced by the definition of $\fvar{Nat}.\opvar{rec}$ is parametric (in a metamathematical sense here) over the representation type of $\tvar{t2}$. In fact, we will demand this for all types constructed by a type constructor other than the one an operator constructor is associated with. This leads us into the concept of conservativity.
%
%
%\subsection{Representational Independence Implies Conservativity}\label{repind}
%In isolation (let us call this language @$\lambda[\fvalCtx_{\mathbf{T}}]$) our definition of natural numbers can be shown to be a strong encoding of the semantics of G\"odel's $\mathbf{T}$. That is, there is a mapping of terms and types into @$\lambda[\fvalCtx_{\mathbf{T}}]$ and an inverse mapping. Indeed, these are the entirely obvious one, given how directly we constructed @$\lambda[\fvalCtx_{\mathbf{T}}]$. 
%
%Typing is preserved by  these mappings, and we can show by a bisimulation argument that the translation produced by @$\lambda[\fvalCtx_{\mathbf{T}}]$ implements the dynamic semantics of G\"odel's $\mathbf{T}$. We must leave full details to an appendix, but the techniques are fundamentally standard when reasoning about compiler correctness, here lifted into the language. A necessary lemma in the proofs, however, is that the value of every translation of an external term of type $\ttype{Nat}{\tunit}$ is a non-negative integer. This is needed to show that the fixpoint computation in the definition of $\fvar{Nat}.\opvar{rec}$ is terminating, as the recursor always does in $\mathbf{T}$. A very similar lemma is needed to show that the translation of every external term of type $\ttype{Nat}{\tunit}$ is equivalent to the translation that would be produced by some combination of applications of $\fvar{Nat}.\opvar{z}$ and $\fvar{Nat}.\opvar{s}$ (corresponding to the canonical forms lemma of $\mathbf{T}$). 
%
%Fortunately, the definitions we have provided thus far admit this lemma. Indeed, the proof is quite straightforward. However, it is also quite precarious because it relies on exhaustive induction over the available operators. As soon as we load another operator or type constructor into the language, these theorems may no longer be \emph{conserved}. We already saw an example where a representationally inconsistent operator caused problems, but the semantics render uses of it ill-typed using the representation schema (in a manner we will make more precise shortly). But there are operator constructors that are representationally consistent, but still cause problems because they violate internal invariants. For example, $$\tops{n1}{\kunit}{i}{a}{\tden{-1}{\ttype{Nat}{\tunit}}}$$
%
%Were this operator constructor introduced inside $\fvar{Nat}$, there is no way for our semantics to distinguish between an incorrect implementation of $\fvar{Nat}$ and a correct implementation of some other type where non-termination of the recursor was the intended behavior (without a separate specification). 
%
%But in any \emph{other} type constructor, this clearly appears to violate the tacit, thus far, abstraction barrier between type constructors. In particular, if we extended @$\lambda[\fvalCtx_{\mathbf{T}}]$ with such a type constructor, let us call it $\fvar{Evil}$, then our crucial lemma relating to non-negativity would not be conserved, and thus, our carefully constructed strong encoding would no longer be a strong encoding. In particular, our canonical forms lemma now has a new case, and the recursor behaves badly. \emph{If we assume an abstraction barrier between type constructors\footnote{A fragment consisting of finitely many type constructors can be easily turned into just one by adding a sum type to the index and case analyzing against it whenever a distinction is necessary.} must exist, we do not need need a specification to see that $\fvar{Evil}.\opvar{n1}$ is violating it}. We can prevent problems by using a technique closely related to the one used to enforce abstraction barriers between modules -- holding the representation type abstract. If $\fvar{Evil}.\opvar{n1}$ cannot know that the representation type of $\ttype{Nat}{\tunit}$ is $\dint$, then its definition is incorrect independent of the specification of $\fvar{Nat}$. 
%\section{Metatheory}\label{theory}
%\subsection{Active Typing}
%\begin{figure}[t]
%\small
%\noindent
%$\fbox{\inferrule{}{\dcheckX{\tauD}{\tau}{\ibar}}}$
%~~~$\Xi ::= \emptyset \pipe \Xi, \fvar{Tycon}$
%\begin{mathpar}
%\inferrule[d-check]{
%	\ddbar{\Xi}{\fvalCtx}{\trepof{\tau}}{\sabs}\\
%%	\eCtxTogCtx{\Xi}{\fvalCtx}{\etCtx}{\gtCtx}\\
%	\checkRC{\emptyctx}{\Xi}{\fvalCtx}{\ibar}{\sabs}
%}{
%	\dcheckX{\tden{\ibar}{\tau}}{\tau}{\ibar}
%}
%
%\inferrule[d-checked]{
%	%\dcheckX{\tden{\ibar}{\tau}}{\tau}{\ibar}
%}{
%	\dcheckX{\tden{\ibar}{\tau}^\checkmark}{\tau}{\ibar}
%}
%\end{mathpar}
%$\fbox{\inferrule{}{\ddbarX{\sbar}{\sabs}}}$
%~~~$\sabs ::= \sabs \rightharpoonup \sabs' \pipe \cdots \pipe \sabsrep{\tau}$
%%~~~$\fvalCtx ::= \fvalCtx_0 \pipe \fvalCtx, \fvalDf$~~~
%\begin{mathpar}
%%\inferrule[pconc-int]{ }{
%%	\ddbarX{\dint}{\dint}
%%}
%\inferrule[abs-trity-parr]{
%	\ddbarX{\sbar_1}{\sabs_1}\\
%	\ddbarX{\sbar_2}{\sabs_2}
%}{
%	\ddbarX{\darrow{\sbar_1}{\sbar_2}}{\darrow{\sabs_1}{\sabs_2}}
%}
%
%\inferrule[abs-trity-cancel]{
%	\ddbarX{\sbar}{\sabs}
%}{
%	\ddbarX{\dup{\titype{\sbar}}}{\sabs}
%}
%
%\inferrule[abs-trity-concrete]{
%	\fvar{Tycon} \in \Xi\\
%	\fval{Tycon}{-}{\taurep}{-} \in \fvalCtx\\\\
%	\tEvalX{\taurep~\tauidx}{\titype{\sbar}}\\
%	\ddbarX{\sbar}{\sabs}
%}{
%	\ddbar{\Xi}{\fvalCtx}{\trepof{\ttype{Tycon}{\tauidx}}}{\sabs}
%}
%
%\inferrule[abs-trity-abstract]{
%	\fvar{Tycon} \notin \Xi
%}{
%	\ddbarX{\trepof{\ttype{Tycon}{\tauidx}}}{\sabsrep{\ttype{Tycon}{\tauidx}}}
%}
%%
%%\inferrule[pconc-prod]{
%%	\ddbarX{\sbar_1}{\sbar_1'}\\
%%	\ddbarX{\sbar_2}{\sbar_2'}
%%}{
%%	\ddbarX{\dpair{\sbar_1}{\sbar_2}}{\dpair{\sbar_1'}{\sbar_2'}}
%%}
%%
%
%\text{\color{gray} (omitted rules follow recursively like \textsc{abs-it-parr})}
%\end{mathpar}
%%$\fbox{\inferrule{}{\eCtxTogCtxX{\etCtx}{\gtCtx}}}$
%%~~~$\gtCtx ::= \emptyctx \pipe \gtCtxX{x}{\sbar}$
%%\begin{mathpar}
%%\inferrule[abs-empty]{ }{
%%	\eCtxTogCtxX{\emptyctx}{\emptyctx}
%%}
%%
%%\inferrule[abs-ctx]{
%%	\eCtxTogCtxX{\etCtx}{\gtCtx}\\
%%	\ddbarX{\trepof{\tau}}{\sbar}
%%}{
%%	\eCtxTogCtxX{\etCtxX{x}{\tau}}{\gtCtxX{x}{\sbar}}
%%}
%%\end{mathpar}
%$\fbox{\inferrule{}{\checkRCX{\ibar}{\sabs}}}$
%~~~$\gtCtx ::= \emptyset \pipe \gtCtx, x \sim \sbar$
%\begin{mathpar}
%\inferrule[abs-i-var]{ }{
%	\checkRC{\gtCtxX{x}{\sbar}}{\Xi}{\fvalCtx}{\evar{x}}{\sbar}
%}
%
%\inferrule[abs-i-fix]{
%	\ddbarX{\sbar}{\sabs}\\
%	\checkRC{\gtCtxX{x}{\sabs}}{\Xi}{\fvalCtx}{\ibar}{\sabs}
%}{
%	\checkRCX{\ifix{x}{\sbar}{\ibar}}{\sbar'}
%}
%%
%%\inferrule[abs-i-lam]{
%%	\ddbarX{\sbar_1}{\sbar_1'}\\\\
%%	\checkRC{\gtCtxX{x}{\sbar_1'}}{\Xi}{\fvalCtx}{\ibar}{\sbar_2}
%%}{
%%	\checkRCX{\ilam{x}{\sbar_1}{\ibar}}{\darrow{\sbar_1'}{\sbar_2}}
%%}
%%
%%\inferrule[abs-i-ap]{
%%	\checkRCX{\ibar_1}{\darrow{\sbar_1}{\sbar_2}}\\\\
%%	\checkRCX{\ibar_2}{\sbar_1}
%%}{
%%	\checkRCX{\iapp{\ibar_1}{\ibar_2}}{\sbar_2}
%%}
%
%\text{\color{gray} (omitted rules follow from a standard statics  similarly)}
%
%%\inferrule[abs int]{ }{
%%	\checkRCX{\iintlit}{\dint}
%%}
%%
%%\inferrule[abs op]{
%%	\checkRCX{\ibar_1}{\dint}\\
%%	\checkRCX{\ibar_2}{\dint}
%%}{
%%	\checkRCX{\iop{\ibar_1}{\ibar_2}}{\dint}
%%}
%%
%%\inferrule[abs pair]{
%%	\checkRCX{\ibar_1}{\sbar_1}\\
%%	\checkRCX{\ibar_2}{\sbar_2}
%%}{
%%	\checkRCX{\ipair{\ibar_1}{\ibar_2}}{\dpair{\sbar_1}{\sbar_2}}
%%}
%%
%%\inferrule[abs fst]{
%%	\checkRCX{\ibar}{\dpair{\sbar_1}{\sbar_2}}
%%}{
%%	\checkRCX{\ifst{\ibar}}{\sbar_1}
%%}
%%
%%\inferrule[abs snd]{
%%	\checkRCX{\ibar}{\dpair{\sbar_1}{\sbar_2}}
%%}{
%%	\checkRCX{\isnd{\ibar}}{\sbar_2}
%%}
%%
%% \inferrule[abs-int-eq]{
%% 	\checkRCX{\ibar_1}{\dint}\\
%% 	\checkRCX{\ibar_2}{\dint}\\\\
%% 	\checkRCX{\ibar_3}{\sbar}\\
%% 	\checkRCX{\ibar_4}{\sbar}
%% }{
%% 	\checkRCX{\iIfEq{\ibar_1}{\ibar_2}{\dint}{\ibar_3}{\ibar_4}}{\sbar}
%% }
%%\inferrule[abs-iterm-inverse]{
%%	\checkRCX{\ibar}{\sbar}
%%}{
%%	\checkRCX{\iup{\titerm{\ibar}}}{\sbar}
%%}
%%
%\inferrule[abs-trans-local-unchecked]{
%	\fvar{Tycon} \in \Xi\\
%	\ddbarX{\trepof{\ttype{Tycon}{\tauidx}}}{\sabs}\\
%	\checkRCX{\ibar}{\sabs}
%}{
%	\checkRCX{\itransof{\tden{\ibar}{\ttype{Tycon}{\tauidx}}}}{\sabs}
%}
%
%\inferrule[abs-trans-local-checked]{
%	\fvar{Tycon} \in \Xi\\
%	\ddbarX{\trepof{\ttype{Tycon}{\tauidx}}}{\sabs}
%}{
%	\checkRCX{\itransof{\tden{\ibar}{\ttype{Tycon}{\tauidx}}^\checkmark}}{\sabs}
%}
%
%\inferrule[abs-trans-foreign-checked]{
%	\fvar{Tycon} \notin \Xi
%}{
%	\checkRCX{\itransof{\tden{\ibar}{\ttype{Tycon}{\tauidx}}^{\checkmark}}}{\sabsrep{\ttype{Tycon}{\tauidx}}}
%}
%\end{mathpar}
%\caption{\small Abstracted internal typing involves typechecking the internal language while holding the representation of all types other than those in the current fragment, listed in the fragment context, $\Xi$, abstract. They key rule is \textsc{abs-trans-foreign-checked}, permitting the translation of a checked derivate that has a type assignment outside the fragment to be used against that type's abstract representation type.}
%\label{ait}
%\end{figure}
%\begin{figure}
%\small
%$\fbox{\inferrule{}{\eraseX{\sbar}{\sigma}}}$
%\begin{mathpar}
%\inferrule[deabs-it-repof]{
%	\fval{Tycon}{-}{\taurep}{-} \in \fvalCtx\\
%	\tEvalX{\taurep~\tauidx}{\titype{\sbar}}\\
%	\eraseX{\sbar}{\sigma}
%}{
%	\eraseX{\trepof{\ttype{Tycon}{\tauidx}}}{\sigma}
%}
%
%\inferrule[deabs-it-cancel]{
%	\eraseX{\sbar}{\sigma}
%}{
%	\eraseX{\dup{\titype{\sbar}}}{\sigma}
%}
%
%\inferrule[deabs-it-parr]{
%	\eraseX{\sbar_1}{\sigma_1}\\
%	\eraseX{\sbar_2}{\sigma_2}
%}{
%	\eraseX{\darrow{\sbar_1}{\sbar_2}}{\darrow{\sigma_1}{\sigma_2}}
%}
%
%\text{\color{gray} (remaining forms follow recursively like \textsc{deabs-it-parr}, omitted)}
%\end{mathpar}
%$\fbox{\inferrule{}{\eraseX{\ibar}{\iota}}}$
%\begin{mathpar}
%\inferrule[deabs-i-var]{ }{\eraseX{x}{x}}
%
%\inferrule[deabs-i-fix]{
%	\eraseX{\sbar}{\sigma}\\
%	\eraseX{\ibar}{\iota}
%}{
%	\eraseX{\ifix{x}{\sbar}{\ibar}}{\ifix{x}{\sigma}{\iota}}
%}
%
%\text{\color{gray} (remaining forms follow recursively like \textsc{deabs-i-fix}, omitted)}
%
%\inferrule[deabs-trans-unchecked]{
%	\eraseX{\ibar}{\iota}
%}{
%	\eraseX{\itransof{\tden{\ibar}{\tau}}}{\iota}
%}
%
%\inferrule[deabs-trans-checked]{
%	\eraseX{\ibar}{\iota}
%}{
%	\eraseX{\itransof{\tden{\ibar}{\tau}^\checkmark}}{\iota}
%}
%\end{mathpar}
%\caption{\small Deabstraction simply recursively eliminates syntactic abstraction barriers and composition-related constructs, and is the final step.}
%\label{deabstraction}
%\end{figure}
%
%allowing us to prove decidability of typechecking for the language as a whole by essentially just citing the termination theorem for the type-level language and proving that the underlying semantics are otherwise well-founded.
%
%\section{Expanded Syntax}\label{expanded-syntax}
%
%\section{Examples}\label{examples}
%\begin{figure}
%\small
%\begin{flalign}
%\small
% & \family{Tuple}{\klist{\kTypeBlur}}{\\
% & \quad \tops{new}{\kunit}{\_}{a}{\tlam{\_}{\kunit}{\tlam{a}{\klist{\kDen}}{\\
% & \quad\quad
% 	\tfold{\tvar{a}}{\tden{{()}}{\ttype{Tuple}{\tnil{\kTypeBlur}}}}{d}{b}{r}{\\
%			& \quad\quad\quad \tfamcase{\ttypeof{\tvar{r}}}{Ntuple}{i}{\\
%			& \quad\quad\quad\quad \tfold{\tvar{b}}{\tden{\itransof{\tvar{d}}}{\ttype{Ntuple}{\tcons{\ttypeof{\tvar{d}}}{\tvar{i}}}}\\& \quad\quad\quad\quad}{\_}{\_}{\_}{\tden{\\& \quad\quad\quad\quad\quad {(\itransof{\tvar{d}},\itransof{\tvar{r}})}}{\ttype{Ntuple}{\tcons{{\ttypeof{\tvar{d}}}}{\tvar{i}}}}}
%			}{\terr}
%	}
% }}};\\
% & \quad \tops{prj}{\kint}{i}{a}{\lambda \tvar{i}:\kint.\lambda \tvar{a}:\klist{\kDen}. 
% 	\tapp{\tvar{arity1}}{\tapp{\tvar{a}}{\tlam{d}{\kDen}{\\
%	& \quad\quad\tfamcase{\ttypeof{\tvar{d}}}{Ntuple}{nl}{\\
% & \quad\quad \tfold{\tvar{nl}}{\terr}{t1}{j}{\_}{\\
% & \quad\quad \tfold{\tvar{j}}{\tifeq{\tvar{i}}{1}{\dint}{\tden{\itransof{d}}{\tvar{t1}}}{\terr}\\&\quad\quad\quad}{\_}{\_}{\_}{\\
% & \quad\quad\quad (\tlam{p}{\kpair{\kDen}{\dint}}{\tifeq{\tsnd{\tvar{p}}}{\tvar{i}}{\dint}{\tfst{\tvar{p}}}{\terr}})\\
% & \quad\quad\quad (\tapp{\tvar{foldl}}{\tapp{\tvar{nl}}{\tapp{(\tden{\tvar{x}}{\tvar{nt}},0)}{
% 	\tlam{r}{\kpair{\kDen}{\dint}}{\tlam{t}{\kTypeBlur}{\tlam{ts}{\klist{\kTypeBlur}}{\\
%	& \quad\quad\quad\quad \tifeq{\tvar{i}}{\tsnd{\tvar{r}}}{\dint}{\tvar{r}}{\tdencase{\tfst{\tvar{r}}}{rx}{\_}{\\
%	& \quad\quad\quad\quad\quad \tifeq{\tvar{i}}{\tsnd{\tvar{r}}+1}{\dint}{\\
%	& \quad\quad\quad\quad\quad\quad \tifeq{\tvar{ts}}{\tnil{\kTypeBlur}}{\klist{\kTypeBlur}}{(\tden{\tvar{rx}}{\tvar{t}},\tvar{i})\\
%	& \quad\quad\quad\quad\quad\quad}{(\tden{\titerm{\ifst{\iup{\tvar{rx}}}}}{\tvar{t}},\tvar{i})}\\	
%	& \quad\quad\quad\quad\quad}{(\tden{\titerm{\isnd{\iup{\tvar{rx}}}}}{\tvar{t}},\tsnd{\tvar{r}}+1)}}{\terr}}}}}
% }}})
% }}{\terr}}}}}
% }\\ & 
%%& \quad \tops{pr}a}{b}{c}{d}
% }{d}{i}{(\tlam{i}{\klist{\kTypeBlur}}{
% 	\\& \quad \quad \tfold{\tvar{i}}{\titype{\dunit}}{s}{j}{r}{
% 		\tfold{\tvar{j}}{\titype{\trepof{\tvar{s}}}}{\_}{\_}{\_}{\\
%		&\quad\quad\quad
% 		\titype{\dpair{\trepof{\tvar{s}}}{\dup{\tvar{r}}}}
% 		}
% 	}
% })}
%\end{flalign}
%\caption{ABC}
%\label{tuple}
%\end{figure}
%
%\section{Limitations and Future Work}\label{limitations}
%\todo{integrate this}Indeed, it is not a stretch to imagine a reasonably faithful implementation of our calculus starting with a pure, total subset of the term language of ML to construct the type-level language. We leave the development of such an implementation (and of bootstrapping type-level and internal languages that are themselves user-defined or extensible) as areas for future work to focus sharply on the theoretical foundations in this work.
%
%Because the input to an operator definition is a list of the recursively determined derivates of the argument in the same typing context as it appears in, our mechanism does not presently permit the definition of operator constructors that bind variables themselves or require a different form of typing judgement (e.g. additional forms of contexts). This is why the $\lambda$ form needs to be built in. It is the only form in the external language that can be used to bind variables. The type constructor $\fvar{Arrow}$ can, however, be defined from within the language and is included in the initial constructor context (it could be written out and included as a ``prelude'' as well). As seen in Fig. \ref{arrow}, it only defines the operator constructor $\opvar{ap}$.
%
%By writing the type assignment and translation logic in this way, as a type-level function, we are taking a rather ``implementation-focused view'' rather than attempting to extract a compiler from a declarative specification of a type system and an operational semantics. That is, @$\lambda$ leaves ultimately to the provider the problem of finding a decidable type assignment algorithm and resolve non-determinism that may arise. Similarly, we do not attempt to turn an operational semantics into a compiler automatically. We  also avoid difficulties with useful error reporting in practice. Simmons' discusses the difficulties of these issues \cite{rjs-princeton}. Composing two dialect specifications directly is more fraught with metatheoretic issues than analyzing two functional programs used together. By looking at the problem as a functional programmer, we can apply intuitions about abstraction barriers more directly. A solution that permits implementations to be extracted from fragment specifications could easily be layered atop this mechanism and inherit these guarantees. Alternatively, extension correctness could be mechanized by encoding these semantics in a proof assistant like Coq and then using its extraction mechanisms to produce a compiler. 
%
%
%The contexts and external term are ``input'' to the judgement, while the type assignment and translation are ``output'' (we briefly discuss potential variants in Sec. \ref{limitations}\todo{remove if not}).
%%At best, they can aim to achieve a separation of concerns, such that each component of an application can be written using a specialized language (as advocated by proponents of the \emph{language-oriented approach} to software development \cite{language-oriented}), but there will still be an {interoperability} problem at the interface between components, as described above, unless an FFI has been explicitly developed for each pair of languages that might interface with each other, or library interfaces are constrained to only use a small subset of common constructs (as suggested in draft guidelines for F\#, for example \cite{fsharp-guidelines}) reducing  expressiveness and safety.
%
%  %We can specify the static semantics of the STLC as follows, using a uniform abstract syntax where type and operator indices are written in braces to emphasize this way of thinking about its structure:
%%\begin{figure}[h]
%%\small
%%\vspace{-8pt}
%%\begin{mathpar}
%%\inferrule[var]{ }{
%%	\jet{\eCtxX{\eCtx}{x}{\tau}}{x}{\tau}
%%}
%%
%%\inferrule[arrow-I]{
%%	\jet{\eCtxX{\eCtx}{x}{\tau}}{e}{\tau'}	
%%}{
%%	\jetX{\lambda[\tau]({x}.e)}{\fvar{Arrow}[(\tau, \tau')]}
%%}
%%
%%\inferrule[arrow-E]{
%%	\jetX{e_1}{\fvar{Arrow}[(\tau, \tau')]}\\
%%	\jetX{e_2}{\tau}
%%}{
%%	\jetX{\texttt{ap}[()](e_1; e_2)}{\tau'}
%%}
%%\end{mathpar}
%%\vspace{-8pt}
%%\caption{Static semantics of the simply typed lambda calculus.}
%%\end{figure}
%
%%\begin{figure}[h]
%%\small
%%\vspace{-8pt}
%%\begin{mathpar}
%%\inferrule[nat-I1]{ }{
%%	\jetX{\texttt{z}[()]()}{\fvar{Nat}[()]}
%%}
%%
%%\inferrule[nat-I2]{
%%	\jetX{e}{\fvar{Nat}[()]}
%%}{
%%	\jetX{\texttt{s}[()](e)}{\fvar{Nat}[()]}
%%}
%%
%%\inferrule[nat-E]{
%%	\jetX{e_1}{\fvar{Nat}[()]}\\
%%	\jetX{e_2}{\tau}\\
%%	\jet{\eCtxX{\eCtxX{\eCtx}{x}{\fvar{Nat}[()]}}{y}{\tau}}{e_3}{\tau}
%%}{
%%	\jetX{\texttt{natrec}[()](e_1; e_2; x.y.e_3)}{\tau}
%%}
%%\end{mathpar}
%%\vspace{-8pt}
%%\caption{Static semantics of natural numbers.}\label{nat-statics}
%%\vspace{-15pt}
%%\end{figure}
%%\begin{figure}[h]
%%\small
%%\begin{mathpar}
%%\inferrule[Prod-I]{
%%	\jetX{e_1}{\tau_1}\\
%%	\jetX{e_2}{\tau_2}
%%}{
%%	\jetX{\texttt{pair}[()](e_1; e_2)}{\fvar{Prod}[(\tau_1, \tau_2)]}
%%}
%%
%%\inferrule[Prod-E1]{
%%	\jetX{e}{\fvar{Prod}[(\tau_1, \tau_2)]}
%%}{
%%	\jetX{\texttt{prl}[()](e)}{\tau_1}
%%}
%%
%%\inferrule[Prod-E2]{
%%	\jetX{e}{\fvar{Prod}[(\tau_1, \tau_2)]}
%%}{
%%	\jetX{\texttt{prr}[()](e)}{\tau_1}
%%}
%%\end{mathpar}
%%\vspace{-8pt}
%%\caption{Static semantics of products}\label{pair-statics}
%%\end{figure}
%
%
%%creating a new language. If this is not practical, the best one can attempt to do is encode the new types in terms of existing types (by a Church encoding, for example). This is generally unsatisfactory -- 
%
%%Languages implemented using these common patterns are central planning by a language designer or design committee. 
%
%%Researchers or domain experts who cannot work around such limitations must develop new standalone languages. In our simple scenario, we may simply copy our implementation of G\"odel's T or even edit it directly (a pernicious technique for implementing a new language where the prior one is overwritten). In a more complex scenario, we may instead employ a tool like a compiler generator or DSL framework \cite{fowler2010domain} that can generate a standalone implementation from declarative specifications of language constructs. Some of these tools allow you to package and reuse these specifications (with the important caveat that not all combinations of constructs are valid and free of conflicts, an important modularity issue that we will return to several times in this paper).
%%
%%The increasing sophistication and ease-of-use of these tools have led many to suggest a {\it language-oriented approach} \cite{journals/stp/Ward94} to software development where different components of an application are written in different languages. Unfortunately, this leads to problems at language boundaries: a library's external interface must only use constructs that can reasonably be expressed in \emph{all possible calling languages}. This can restrict domain-specific languages by, for example, precluding constructs that rely on statically-checked invariants stronger than those their underlying representation in a common target language normally supports. At best, constructs like these can be exposed by generating a wrapper where run-time checks have been inserted to guarantee necessary invariants. This compromises both verifiability and performance and requires the development of an interoperability layer for every DSL. Moreover, library clients must work with verbose and unnatural ``glue code'' when interfacing across languages, defeating the primary purpose of high-level programming languages: hiding the low-level details from the end-users of abstractions. We diagram this fundamental \emph{compatibility problem} in Figure \ref{approaches}(a).
%%\begin{figure*}
%%\begin{center}
%%\includegraphics[scale=0.5]{approaches.pdf}
%%\end{center}
%%\vspace{-20px}
%%\caption{\small (a) With a language-oriented approach, novel constructs are packaged into separate languages. Users can only safely and naturally call into languages consisting of common constructs (often only the common target language, such as C or Java bytecode). (b) With a language-internal extensibility approach, there is one system providing a common internal language, where additional primitive constructs that strictly strengthen its static guarantees or perform specialized code generation are specified and distributed within libraries. \label{approaches}}
%%\end{figure*}
%
%%As a result, domain-specific languages and new general-purpose abstractions alike have experienced relatively slow adoption in practice.
%%
%%Porting large codebases to new languages is difficult, and the dominant programming languages innovate slowly, so programming language.
%%
%%More specifically, such languages are neither \emph{internally extensible} because the language itself exposes only natural numbers and functions to its users, nor are they \emph{externally extensible} because no new behaviors can be added to the language's  implementation in a separate module from the one containing the initial implementation.
%
%%This is the essence of a monolithic language implementation: it is impossible for anyone to modularly extend languages defined in this way. 
%
%%An extensible programming language could address these problems by providing a language-integrated mechanism for introducing new type and operator constructors and implementing their associated static and dynamic semantics directly. 
%%%Library developers need only consider which abstractions are most appropriate for their domain, without also considering whether these constructs can be exposed using abstractions appropriate to the domains of client code. Clients can simply import any necessary constructs when using a library that relies on them, preserving safety and ease-of-use without the use of  wrappers and glue code. We show this competing approach in Figure \ref{approaches}(b).
%%%Researchers and domain experts thus gain the ability to distribute new ideas for evaluation to a broader development community without requiring the approval of maintainers of mainstream languages, large-scale porting of code or explicit interoperability layers. 
%%But, as mentioned in Section \ref{language-integrated-approaches}, some significant challenges must be addressed before such a mechanism can be relied upon. The desire for expressiveness must be balanced against  concerns about maintaining various safety properties in the presence of arbitrary combinations of user-defined  extensions to the language's core semantics. The mechanism must ensure that desirable \emph{metatheoretic properties} (e.g. type safety, decidability) of the language are maintained by extensions. Because multiple independently developed extensions might be used within one program, the mechanism must further guarantee their \emph{non-interference}. These are the issues we seek to address in this work.
%
%%When designing and implementing a new abstraction, experts typically begin by attempting to define new constructs in terms of existing language constructs.
%%This approach is often effective because modern {general-purpose} abstraction mechanisms, like inductive datatypes and object systems, are highly expressive.  
%%For example, the Delite framework leverages Scala's powerful general-purpose mechanisms to enable a number of useful \emph{embedded domain-specific languages} \cite{delite}.
%%Unfortunately, there remain some situations of interest where general-purpose abstractions fall short. 
%%For instance, it is difficult to adequately encode advanced type systems in terms of the simpler rules that govern general-purpose abstractions (e.g. reasoning about units of measure requires built-in language support in F\# \cite{conf/cefp/Kennedy09}). 
%%Even if a full encoding is possible, it may not be useful if it is overly verbose or unnatural, or if the error messages are overly abstract. For example, regular expressions encoded using inductive datatypes are quite verbose, so most functional languages support them via strings, which is less safe. 
%%Finally, general-purpose abstractions are implemented in a uniform manner, meaning domain-specific heuristics cannot be applied to eliminate overhead or perform optimizations, and implementations designed for typical application workloads may not be satisfactory in parts of a program where performance is a key criteria, such as when targeting heterogeneous hardware platforms (e.g. programmable GPUs) and distributed computing resources.
%%
%%% are at times impossible or impractical. In our example of adding products or sums to Godel's T, although Church encodings are possible \cite{pfpl}, they require a reasonable level of creativity\footnote{Anecdotally, Church encodings are among the more difficult-to-explain topics covered in our undergraduate programming languages course.}. Moreover, they will not offer the same static safety guarantees as a primitive encoding, they are more verbose and they will incur performance overhead by their use of closures rather than a more direct representation. This is not only a problem for simple languages like Godel's T. Several Haskell-based embedded DSLs have also needed to make significant compromises at times \cite{haskellDSLs}. {\color{red} examples? Scala?}
%%
%%%creating a new language. If this is not practical, the best one can attempt to do is encode the new types in terms of existing types (by a Church encoding, for example). This is generally unsatisfactory -- 
%%
%%%Languages implemented using these common patterns are central planning by a language designer or design committee. 
%%
%%Researchers or domain experts who run into situations like these, where more direct control over a language's semantics and implementation are needed, have little choice today but to realize new abstractions by creating a new language in some way. They might develop a new standalone language from scratch, modify an implementation of an existing language, or use tools like compiler generators, DSL frameworks and language workbenches \cite{fowler2010domain}. 
%%%In our simple scenario, we may simply fork our implementation of Godel's T or even edit it directly (a pernicious technique for implementing a new language where the prior one is overwritten). 
%%%In a more complex scenario, we may instead employ a tool like a compiler generator or DSL framework \cite{fowler2010domain} that can generate a standalone implementation from declarative specifications of language constructs. Some of these tools allow you to package and reuse these specifications (with the important caveat that not all combinations of constructs are valid and free of conflicts, an important modularity issue that we will return to several times in this paper).
%%The increasing sophistication and ease-of-use of these tools have led to calls for a {\it language-oriented approach} to software development, where different components of an application are written in different specialized languages \cite{journals/stp/Ward94}. Indeed, a number of software ecosystems are now designed explicitly to support many different languages, both general-purpose and domain-specific, atop a common intermediate language. The Java virtual machine (JVM), the Common Language Infrastructure (CLI) and LLVM are prominent examples of such ecosystems.
%%
%%Unfortunately, this leads to a critical problem at language boundaries: a library's external interface must only use constructs that can reasonably be expressed in \emph{all possible client languages}. This discourages languages from including constructs that rely on statically-checked invariants stronger than those supported by their underlying implementation in the common intermediate language. At best, constructs like these can be exposed by generating a wrapper where run-time checks have been inserted to guarantee these invariants. This compromises both verifiability and performance. %Forrequires the development of an interoperability layer for every pair of DSLs. 
%%Moreover, this approach exposes the internals of an implementation to clients, making the abstraction awkward to work with and causing code breakage when implementation details change. This defeats a primary purpose of high-level programming languages: hiding low-level details from clients of an abstraction. We diagram this fundamental \emph{interoperability problem} in Figure \ref{approaches}(a). 
%%%As an example, F\#'s type system prevents \lstinline{null} values from occurring within data structures, but because it's type system is not available when calling into F\# code from another language, like C\#, run-time null checks must still be included in the implementation.
%%%\begin{figure*}[t]
%%%\vspace{-15px}
%%%\includegraphics[scale=0.415]{approaches.pdf}
%%%\caption{(a) With the language-oriented approach, different primitive abstractions are packaged into separate languages that extend and target a common intermediate language (e.g. JVM bytecode). Users can only interface with libraries written in another language via the constructs in the common language, causing \emph{interoperability problems}. (b) With the language-internal approach, the semantics of new abstractions (i.e. the logic that governs typechecking and translation to a fixed {internal language}, here labeled \texttt{I}) can be implemented directly within so-called \emph{active libraries}. Clients can import and use these abstractions directly whenever needed.}
%%%\label{approaches}
%%%\end{figure*}
%%
%%%As a result, domain-specific languages and new general-purpose abstractions alike have experienced relatively slow adoption in practice.
%%%
%%%Porting large codebases to new languages is difficult, and the dominant programming languages innovate slowly, so programming language.
%%%
%%%More specifically, such languages are neither \emph{internally extensible} because the language itself exposes only natural numbers and functions to its users, nor are they \emph{externally extensible} because no new behaviors can be added to the language's  implementation in a separate module from the one containing the initial implementation.
%%
%%%This is the essence of a monolithic language implementation: it is impossible for anyone to modularly extend languages defined in this way. 
%%
%%%Programming languages are typically designed around a monolithic collection of primitive type families and operators. Consider, as a simple example, Godel's T \cite{pfpl}, a typed lambda calculus with recursion on primitive natural numbers\todo{add statics to Appendix A}. Although a language designer may casually speak of ``extending Godel's T with primitive product and sum types'', adding these type families and associated operators to this kind of language from within is impossible. That is, Godel's T is not \emph{internally extensible}.
%%
%%\emph{Internally-extensible programming languages} promise to avoid these problems by providing researchers and domain experts with a mechanism for implementing the semantics of new primitive constructs directly within libraries.
%%%Developers of libraries need only determine whether they are appropriate for their domain, without also considering whether these constructs can be exposed in terms of abstractions appropriate to client code. 
%%As a result, clients can granularly import any necessary primitive constructs when using code that relies on them, and thereby achieve full safety, ease-of-use and performance. Providers of components thus need only consider whether primitives that they use are appropriate for their domain, without also considering whether their code might be used in a context where these primitives are not otherwise appropriate. Libraries containing logic that is invoked at compile-time, as extension logic would be, have been called \emph{active libraries} \cite{activelibraries}. We adopt this terminology and diagram this competing approach in Figure \ref{approaches}(b).
%%
%%%Researchers and domain experts thus gain the ability to distribute new ideas for evaluation to a broader development community without requiring the approval of maintainers of mainstream languages, large-scale porting of code or explicit interoperability layers. 
%%
%%For a language-internal extension mechanism to be feasible, however, it must achieve expressiveness while also ensuring that extensions cannot compromise the safety properties of the language and its tools, nor interfere with one another. That is, extensions cannot simply be permitted to add arbitrary logic to the type system or compiler, because this would make it possible to break  type safety, decidability or adequacy theorems that are critical to the operation of the language, the compiler or other extensions. We review some previous attempts at language extensibility, and highlight how they do not adequately achieve both safety and expressiveness, in Section \ref{related-work}.
%%%{\color{red} transition here} Correctness properties of an extension itself should be modularly verifiable, so that its users can rely on it for verifying and compiling their own code. The mechanism must also ensure that desirable metatheoretic properties and global safety guarantees of the language cannot be weakened by extensions. And with multiple independently-developed extensions used at once, the mechanism must further guarantee that they cannot interfere with one another. 
%%
%%In this paper, we introduce a language-internal extensibility mechanism called \emph{active typechecking and translation} (AT\&T) that allows developers to introduce and implement the logic governing new primitive type and operator families from within libraries. 
%%We argue that this can be accomplished by enriching the type-level language, rather than introducing a separate metalanguage into the system. 
%%To make this proposal concrete, we begin by introducing a simple core calculus, called \atlam~(for the ``actively-typed lambda calculus''), in Section \ref{atlam}. 
%%This calculus uses type-level computation of higher kind, along with techniques borrowed from the typed compilation literature and a form of type abstraction that ensures that the implementation details of an extension are not externally visible to guarantee the safety of the language, the decidability of typechecking and compilation and non-interference of extensions, as we outline in Section \ref{safety}.
%%In Section \ref{examples}, we suggest that despite these constraints, this mechanism is expressive enough to admit, within libraries, a number of general-purpose and domain-specific abstractions that normally require built-in language support. 
%%
%%Our core calculus uses a uniform abstract syntax for primitive operators to simplify our presentation and analysis, but this syntax is too verbose to be practical. Thus, we begin this section by showing how a key design choice made in the calculus -- to associate operators with type families, forming what we call \emph{active type families} -- supports a novel type-directed desugaring mechanism that permits the use of conventional concrete syntax for language extensions. 
%
%% Our choice of a simply-typed, simply-kinded calculus where expressions are given meaning by translation to a simply-typed internal language appears to occupy a ``sweet spot'' in the design space, and relates closely to how simply-typed functional languages like  ML and Haskell are specified and implemented today. In Section \ref{design}, we briefly discuss other points in the design space of actively-typed languages and describe the sorts of abstractions that the mechanism as we have introduced it is not capable of expressing, suggesting several directions for future research. We conclude with a discussion of related work in Section \ref{related-work}.
%%dependently-typed and object-oriented type-level languages, as well as the constraints governing the design of the internal language.  
%%We will also note how object-oriented techniques may also be suit, because of a fundamental connection to the \emph{expression problem} \cite{expression-problem}.
%
%%specify new typechecking rules and translation logic from within libraries. The AT\&T mechanism utilizes type-level computation of higher kind and integrates typed compilation techniques into the language to provide strong safety guarantees, while remaining straightforward and expressive.
%
%%AT\&T is general with respect to many choices about the type-level language, the typed internal language and syntax. Choices along these dimensions can affect both expressiveness and ease-of-use. We will begin in Sec. 2 by introducing a minimal system called $@\lambda$ (the ``actively-typed lambda calculus'') that distills the essence of the mechanism in a simply-typed, simply-kinded setting. This will allow us to fully and concisely formalize the language and compiler and give several key safety theorems. We will then continue in Sec. 3 by discussing variants of this mechanism based on other basic paradigms, considering dependently-typed functional languages and object-oriented languages, discussing trade-offs between expressivity and safety when doing so. We have developed a simple prototype called Ace and have used it to develop a number of full-scale language extensions as libraries. We will briefly discuss this language and these extensions in Sec. 4.
%
%%We note at the outset that AT\&T focuses on extending the static semantics of languages with fixed, though flexible, syntax. Language-internal syntax extension mechanisms have been developed in the past (e.g. SugarJ \cite{sugarj}) but they have also suffered from safety problems because grammar composition is not always safe when done in an  unconstrained manner. Constrained approaches that provide stronger safety guarantees have recently been outlined (e.g. Wyvern \cite{globaldsl13}) but we will leave integration of syntax extensions with semantic extensions as future work.
%%\section{From Extensible Compilers to Extensible Languages}\label{evolution}
%%\begin{figure}[t]
%%\small
%%$$\begin{array}{rccl}	
%%\textbf{programs} & \rho & ::= & \pfam{\familyDf}{\progsort}  \\
%% & & \pipe &  \pdef{t}{\kappa}{\tau}{\progsort} \pipe e\\
%%\text{primitive ops}		&	\theta	&	::= &	\tops{op}{\kappaidx}{i}{a}{\tau} \pipe 
%%												\topp{\theta}{\theta}\\
%%\\
%%\textbf{external terms} 				&	e	&	::=	&	\evar{x} \pipe 
%%														\elam{\evar{x}}{\tau}{e} \pipe 
%%														\eop{Fam}{op}{
%%															\taui
%%														}{
%%  												    		\splat{e}{1}{n}
%%														} \\
%%									& 		&		& 	\\
%%							
%%\hspace{-5pt}\textbf{type-level terms} 	& \tau 	& ::= 	& 	\tvar{t} \pipe 
%%														\tlam{t}{\kappa}{\tau} \pipe 
%%														\tapp{\tau_1}{\tau_2}\pipe
%%														\iintlit \pipe \iop{\tau_1}{\tau_2} \pipe \tstr{str} \\
%%									    & &  \pipe & 
%%														\tnil{\kappa} \pipe \tcons{\tau_1}{\tau_2} \pipe 
%%									                     \tfold{\tau_1}{\tau_2}{h}{t}{r}{\tau_3}
%%														\\
%%												
%%	 			& 		& \pipe	& 	 \tunit \pipe 
%%														\tpair{\tau_{1}}{\tau_{2}} \pipe 
%%														\tfst{\tau} \pipe 
%%														\tsnd{\tau} 
%%														\\
%%\text{equality}  & & \pipe & 					\tifeq{\tau_{1}}{\tau_{2}}{\kappa}{\tau_{3}}{\tau_{4}} 
%%														\\														
%%
%%\text{types} 						& 		& \pipe	& 	\ttypestd \\
%% & & \pipe &  \tfamcase{\tau}{Tycon}{x}{\tau_1}{\tau_2}\\
%%						%				& & \pipe & \tfamcase{\tau}{Fam}{x}{\tau_1}{\tau_2}\\
%%																								
%%\text{denotations} 				& 		 & 	\pipe	&	\tden{\tauiterm}{\tautype} \pipe \terr \\
%% & & \pipe &  \tdencase{\tau}{x}{t}{\tau_1}{\tau_2}\\
%% %& & \pipe & 
%%%														\tdencase{\tau}{y}{x}{\tau_1}{\tau_2}
%%%														 \\
%%
%%\text{reified IL}		&		&	\pipe	&	\titerm{\iota} \pipe \titype{\sigma} \\
%%
%%												\\
%%\textbf{kinds} 					& \kappa	&	::=	&	\karrow{\kappa_1}{\kappa_2} \pipe \klist{\kappa} \pipe \dint \pipe
%%											    \kstr \pipe
%%												\kunit \pipe 
%%												\kpair{\kappa_{1}}{\kappa_{2}} \\
%%	& & \pipe &  
%%												\kTypeBlur \pipe \kDen \pipe 
%%												\kIType \pipe \kITerm
%%												\\
%%\\												
%%%\textbf{ops signature}			& \Theta	&	::=	&	\kOpEmpty \pipe \kOp{\Theta}{op}{\kappai}\\
%%%											 							&		&		&	\\
%%\textbf{internal terms} 				& 	\iota	&	::=	&	\evar{x} \pipe 
%%												\ilam{\evar{x}}{\sigma}{\iota} \pipe 
%%												\iapp{\iota_{1}}{\iota_{2}} \pipe
%%												\ifix{\evar{f}}{\sigma}{\iota} \\
%%									& & \pipe & 
%%												\ipair{\iota_{1}}{\iota_{2}} \pipe 
%%												\ifst{\iota} \pipe
%%												\isnd{\iota}  
%%												\\
%%							& 		& 	\pipe	& 
%%												\iintlit \pipe \iop{\iota_{1}}{\iota_{2}} \pipe \iIfEq{\iota_{1}}{\iota_{2}}{\dint}{\iota_{3}}{\iota_{4}}  
%%												\\
%% &  & \pipe & \tvalof{\tau_1}{\tau_2} \pipe \iup{\tau} \\
%%%\text{deabstracted}& \iota & ::= & \mathcal{G}[\iota, \sigma]\\
%%\textbf{internal types}			&	\sigma	&	::=	&    \darrow{\sigma_1}{\sigma_2} \pipe
%%												\dint \pipe
%%												\dpair{\sigma_1}{\sigma_2} \pipe
%%												\trepof{\tau} \pipe \dup{\tau}\\
%%\end{array}$$
%%%\vspace{-10pt}
%%\caption{\small Syntax of \atlam. Variables $x$ are used in expressions and internal terms and are distinct from type-level variables, $\tvar{t}$. Names $\fvar{Fam}$ are type family names (we assume that globally unique type family names can be generated by some external mechanism) and $\opvar{op}$ are operator family names. $\tstr{str}$ denotes string literals, $\iintlit$ denotes integer literals and $\oplus$ stands for binary operations over integers. %The productions related to the internal language are written using generators $\mathcal{G}$ and $\mathcal{S}$ to avoid duplicating the syntax of common terms.
%%\label{grammar}}
%%\end{figure}
%%\begin{figure}[t]
%%\small
%%\begin{flalign}
%%\label{natfam}&\family{Nat}{\kunit}{\\
%%\label{z}&\quad\tops{z}{\kunit}{i}{a}{
%%	\tapp{\tapp{\tvar{empty}}{\tvar{a}}}{
%%		\tden{\titerm{0}}{\ttype{Nat}{\tunit}}
%%	}};\\
%%\label{s}&\quad\tops{s}{\kunit}{i}{a}{
%%	\tapp{
%%		\tapp{\tvar{pop\_final}}{\tvar{a}}}{\tlam{x}{\kITerm}{\tlam{t}{\kTypeBlur}{
%%			\\
%%			&\quad\quad\tapp{\tapp{\tapp{\tvar{check\_type}}{\tvar{t}}}{\ttype{Nat}{\tunit}}}{
%%				\tden{\titerm{\iup{\tvar{x}}+1}}{\ttype{Nat}{\tunit}}
%%			}
%%		}}
%%		}
%%	};\\
%%\label{rec}&\quad\tops{rec}{\kunit}{i}{a}{
%%	\tapp{\tapp{\tvar{pop}}{\tvar{a}}}{\tlam{x1}{\kITerm}{\tlam{t1}{\kTypeBlur}{\tlam{a}{\klist{\kDen}}{\\
%%	&\quad\quad\tapp{\tapp{\tvar{pop}}{\tvar{a}}}{\tlam{x2}{\kITerm}{\tlam{t2}{\kTypeBlur}{\tlam{a}{\klist{\kDen}}{\\
%%	&\quad\quad \tapp{\tapp{\tvar{pop\_final}}{\tvar{a}}}{\tlam{x3}{\kITerm}{\tlam{t3}{\kTypeBlur}{\\
%%	&\quad\quad \tapp{\tapp{\tapp{\tvar{check\_type}}{\tvar{t1}}}{\ttype{Nat}{\tunit}}}{(\\
%%	&\quad\quad \tapp{\tapp{\tapp{\tvar{check\_type}}{\tvar{t3}}}{\ttype{Arrow}{(\ttype{Nat}{\tunit},\ttype{Arrow}{(\tvar{t2},\tvar{t2})})}}}{\\
%%		\label{fix}&\quad\quad \tden{\titerm{\iapp{(\ifix{f}{\darrow{\dint}{\trepof{\tvar{t2}}}}{\ilam{x}{\dint}{\\
%%		\label{lastop}&\quad\quad\quad \iIfEq{x}{0}{\dint}{\iup{\tvar{x2}}}{\iapp{\iapp{\iup{\tvar{x3}}}{(x-1)}}{(\iapp{f}{(x-1)})}}}})}{\iup{\tvar{x1}}}}}{\tvar{t2}}
%%	)}}
%%	}}}}}}}}}}}}
%%\\
%%&}{XXX}{i}{\titype{\dint}};\\
%%\label{nattype}&\pdef{nat}{\kTypeBlur}{\ttype{Nat}{\tunit}}{\\
%%\label{plus}&(\elam{plus}{\ttype{Arrow}{(\tvar{nat}, \ttype{Arrow}{(\tvar{nat}, \tvar{nat}))}}}{
%%	\elam{two}{\tvar{nat}}{\\
%%\label{ap}	&\quad\quad \eopapp{\eopapp{plus}{two}}{two}}})
%%}\\
%%\label{add}&\quad (\elam{x}{\tvar{nat}}{\elam{y}{\tvar{nat}}{
%%	\eop{Nat}{rec}{\tunit}{x; y; \elam{p}{\tvar{nat}}{\elam{r}{\tvar{nat}}{
%%	\eop{Nat}{s}{\tunit}{r}
%%	}}}}})\\
%%\label{two}&\quad  \eop{Nat}{s}{\tunit}{\eop{Nat}{s}{\tunit}{\eop{Nat}{z}{\tunit}{ }}}
%%%{\\&\elet{two}{\tvar{nat}}{\eop{Nat}{s}{\tunit}{\eop{Nat}{s}{\tunit}{\eop{Nat}{z}{\tunit}{ }}}}{\eapp{\eapp{plus}{two}}{two}}}
%%%}
%%%\elam{x}{\tvar{nat}}{\elam{y}{\tvar{nat}}{\\
%%	%&\quad \eop{Nat}{rec}{\tunit}{x; y; \elam{p}{\tvar{nat}}{\elam{r}{\tvar{nat}}{
%%	%\eop{Nat}{s}{\tunit}{r}
%%	%}}}}
%%\end{flalign}
%%\caption{G\"odel's T in \atlam, used to calculate 2+2. Helper functions for working with lists ($\tvar{empty}$, $\tvar{pop}$, $\tvar{pop\_final}$) and types ($\tvar{check\_type}$),  described below, are given in the appendix. We will refer to this program as $\rho_{\text{nat}}$ in the text.}
%%\label{example}
%%\end{figure}
%%
%%To understand the genesis of our internal extension mechanism, it is helpful to begin by considering why most implementations of programming languages cannot even be  externally extended. 
%%Let us consider, as a simple example, an implementation of G\"odel's T, a typed lambda calculus with recursion on primitive natural numbers (see Appendix). 
%%A compiler for this language written using a functional language will invariably represent the primitive type families and operators using {closed} inductive datatypes. 
%%For example, a simple implementation in Standard ML may be based around these datatypes:
%%\begin{lstlisting}
%%  datatype Type = Nat | Arrow of Type * Type
%%  datatype Exp = Var of var 
%%               | Lam of var * Type * Exp | Ap of Exp * Exp 
%%               | Z | S of Exp | Natrec of Exp * Exp * Exp
%%\end{lstlisting}
%%
%%The logic governing typechecking and translation to a suitable intermediate language (for subsequent optimization and compilation by some back-end) will proceed by exhaustive case analysis over the constructors of \lstinline{Exp}.
%%
%%In an object-oriented implementation of Godel's T, we might instead encode types and operators as subclasses of abstract classes \lstinline{Type} and \lstinline{Exp}. Typechecking and translation will proceed by the ubiquitous \emph{visitor pattern}  by dispatching against a fixed collection of {known} subclasses of \lstinline{Exp}. 
%%
%%In either case, we encounter the same basic issue: there is no way to modularly add new primitive type families and operators and implement their associated typechecking and translation logic. 
%%%This issue is related to the widely-discussed \emph{expression problem} (in a restricted sense -- we do not consider adding new functions beyond typechecking and translation here, only adding logic to these) \cite{wadler-expression}.
%%
%%A number of language mechanisms have been proposed that allow new cases to be added to datatypes and the functions that operate over them in a modular manner. 
%%In functional languages, we might use \emph{open datatypes}. For example, if we wish to extend G\"odel's T with product types and we have written our compiler in a language supporting open inductive datatypes, it might be possible to add new cases like this: 
%%\begin{lstlisting}
%%  newcase Prod of Type * Type extends Type
%%  newcase Pair of Exp * Exp extends Exp    (* Intro *)
%%  newcase PrL of Exp extends Exp           (* Elim Left *)
%%  newcase PrR of Exp extends Exp           (* Elim Right *)
%%\end{lstlisting}
%%
%%The logic for functionality like typechecking and translation could then be implemented for only these new cases. For example, the \lstinline{typeof} function that assigns a type to an expression could be extended like so:
%%\begin{lstlisting}
%%  typeof PrL(e) = case typeof e of 
%%      Prod(t1, _) => t1 
%%    | _ => raise TypeError("<appropriate error message>")
%%\end{lstlisting}
%%
%%If we allowed users to define new modules containing definitions like these and link them into our compiler, we will have succeeded in creating an externally-extensible compiler, albeit one where safety is not guaranteed (we will return to this point shortly). We have not, however, created an extensible programming language, for two reasons. First, compiler extensions are distributed and activated separately from libraries, so dependencies become more difficult to manage. Second, other compilers for the same language will not necessarily support the same extensions. 
%%If our newly-introduced constructs are exposed at a library's  interface boundary, clients using different compilers face the same problems with interoperability that those using different languages face. That is, {extending a language by extending a single compiler for it is morally equivalent to creating a new language}. Several prominent language ecosystems today are in a state where a prominent compiler has introduced or enabled the introduction of extensions that many libraries have come to rely on, including the Glasgow Haskell Compiler, SML/NJ and the GNU compilers for C and C++.
%%
%%A more appropriate and useful place for extensions like this is directly within libraries, alongside abstractions that can be adequately implemented in terms of existing primitive abstractions. To enable this, the language must allow for the introduction new primitive type families, like \lstinline{Prod}, operators, like \lstinline{Pair}, \lstinline{PrL} and \lstinline{PrR}, and associated typechecking and translation logic. When encountering these new operators in expressions, the compiler must effectively  hand control over typechecking and translation to the appropriate user-defined logic. Because this mechanism is {language-internal}, all compilers must support it to satisfy the language specification.
%%
%%Statically-typed languages typically make a distinction between \emph{expressions}, which describe run-time computations, and type-level constructs like types, type aliases and datatype declarations. The design described above suggests we may now need to add another layer to our language, an {extension language}, where extensions can be declared and implemented. In fact, we will show that \textbf{the most natural place for type system extensions is within the type-level language}. The intuition is that extensions to a statically-typed language's semantics will need to manipulate types as values at compile-time. Many languages already allow users to write type-level functions for various reasons, effectively supporting this notion of types as values at compile-time (see Sec. \ref{related-work} for examples). The type-level language is often constrained by its own type system (where the types of type-level values are called \emph{kinds} for clarity) that prevents type-level functions from causing problems during compilation. This is precisely the structure that a distinct extension layer would have, and so it is quite natural to unify the two, as we will show in this work.
%%
%%\section{\atlam}\label{atlam}
%%In this section, we will develop a core calculus, called @$\lambda$ for the ``actively-typed lambda calculus'', by way of a semantics and a simple example, and discuss how it addresses the safety concerns that arise. 
%%\subsection{Overview}
%%The grammar of \atlam~is shown in Figure \ref{grammar}. 
%%A program, $\rho$, consists of a series of declarations followed by an expression. Declarations can be either bindings of type-level terms to type-level variables using \textsf{def} or a primitive type family declared using \textsf{family}, which can contain implementations of one or more operator families, $\theta$. Expressions, $e$, can be either variables, lambdas, or applications of operators.
%%%, and are ultimately given meaning by translation to a typed internal language, with terms $\iota$ and types $\sigma$. 
%%%This language has been chosen, for simplicity, to be a variant of Plotkin's PCF with primitive integers and products, but in practice would include other constructs consistent with its role as a high-level intermediate language.
%%
%%The language is structured as a simply-typed lambda calculus with simply-kinded type-level computation. Kinds, $\kappa$, classify type-level terms, $\tau$. 
%%Types are type-level values of kind $\star$ (following System $F_{\omega}$) and classify expressions. The type-level language also includes other kinds of terms: type-level functions, lists (required by our mechanism), integers, strings and products for the sake of our examples (see Sec. \ref{safety} for a discussion on other acceptable kinds of type-level data) and constructs for developing extensions -- denotations and reified internal terms and types -- which we will discuss in the sections below. 
%%
%%All expressions are given meaning by translation to a typed internal language. This language has been chosen, for simplicity, to be a variant of Plotkin's PCF with primitive integers and products, but in practice would include other constructs consistent with its role as a high-level intermediate language. The grammar of internal terms, $\iota$, and internal types, $\sigma$, also includes special forms containing type-level terms; these are used for developing extensions and during compilation and will be erased before compilation ends.
%%\subsection{Example: G\"odel's T as an Active Type Family}
%%To make our explanation of each of the constructs in the calculus concrete, we will work through an example showing how to introduce primitive natural numbers with bounded recursion in the style of G\"odel's T \cite{pfpl}. These will be implemented internally as integers (that is, internal terms of internal type $\dint$). Figure \ref{example} shows how to define the indexed type family $\fvar{Nat}$. This family contains only one type, written $\ttype{Nat}{\tunit}$, which we alias on line \ref{nattype} by defining the type-level variable $\tvar{nat}$. We define the typechecking and translation logic for the operators associated with this family (\opvar{z}, \opvar{s} and \opvar{rec}) on lines \ref{z}-\ref{lastop} and use these to define a $plus$ function on line \ref{add} and compute 2+2. We will refer back to this figure as we describe each construct below.
%%
%%\subsection{Indexed Type Families and Types}\label{families}
%%The syntactic form $\familyDf$ declares a new primitive type family named $\fvar{Fam}$ indexed by type-level values of kind $\kappaidx$ with representation schema $\tvar{i}.\taurep$ and operators $\theta$. The purpose of the representation schema and of associating of operators directly with types will be explained below. 
%%
%%Declaring a type family in this way is a language-internal analog to adding a new constructor to the compiler-internal datatype \lstinline{Type}, as suggested in Sec. \ref{evolution}. 
%%The index represents the data associated with this constructor. A type (that is, a type-level term of kind $\kTypeBlur$) is constructed by naming a family in scope and providing a type-level term of the appropriate kind as an index. A base type like $\tvar{nat}$ can be thought of as being the only type in the family $\fvar{Nat}$ trivially indexed by the unit value, of kind $\kunit$, while families like $\fvar{Ntuple}$ might be indexed by a list of types, having kind $\klist{\kTypeBlur}$. 
%% For example, $\ttype{Ntuple}{\tcons{\tvar{nat}}{\tcons{\tvar{nat}}{\tnil{\kTypeBlur}}}}$ might be the type of a pair of natural numbers. Given a type, its family can be \textsf{case} analyzed to extract the value of its index. It is important that type equality be decidable, so only kinds for which equivalence coincides with syntactic equality can be used as type family indices. The main  consequence of this restriction is that indices cannot contain type-level functions.
%%
%%\subsection{Representations and Representation Schemas}
%%As we will discuss further below, it is important that all expressions classified by a type compile to consistently-typed internal terms. For this reason, we require that every type have a single internal type associated with it, called its \emph{representation}. This is computed by substituting the type index for the bound variable $\tvar{i}$ in the term $\taurep$, called the \emph{representation schema} of the type family, and evaluating to a value representing an internal type. Internal types, $\sigma$, are reified as type-level terms of kind $\kIType$ using the introductory form $\titype{\sigma}$.
%%
%%\subsection{Indexed Operator Families and Denotations}\label{operators}
%%Type families are also equiped with a collection of primitive operator families. An operator family named $\opvar{op}$ is declared using the form $\tops{op}{\kappaidx}{i}{a}{\tauop}$. Like type families, operator families are indexed by values of some kind, $\kappaidx$, but because operators are not first-class type-level values in our calculus, there are no equality restrictions. In the example in Fig. \ref{example}, all the operators are trivially indexed by the kind $\kunit$, so each family only contains one operator. However, a type family like $\fvar{Ntuple}$ would be equipped with a family of projection operators, $\opvar{pr}$, indexed by a position (e.g. an integer in our calculus). A family implementing record types or object types might have a similar operator indexed by a type-level string representing the field being accessed (see Sec. \ref{examples}).
%%
%%To apply an operator, the grammar provides a uniform form of expression: $\eop{Fam}{op}{\tauidx}{e_1; \ldots; e_n}$, where $n \geq 0$. The typechecking and translation of an expression of this form is controlled by the term $\tauop$ in the operator's declaration. This term must evaluate to a \emph{denotation}, which is a type-level value of kind $\kDen$, when given the operator index, $\tauidx$, and a list constructed from the denotations recursively assigned to each argument, $e_1$ through $e_n$. 
%%
%%There are two forms of denotations that an expression can be assigned. A \emph{valid denotation} has the form $\tden{\tauiterm}{\tautype}$, where $\tauiterm$ is the \emph{translation} of the expression to an internal term and $\tautype$ is the type it has been assigned. Internal terms are represented as type-level terms using the form $\titerm{\iota}$, and have kind $\kITerm$ (similar to $\kIType$). %It is important to note that there are no elimination form for reified terms or types.
%%%(indeed, in \atlam, terms are never  syntactically analyzed directly by extensions, unlike macro systems and other forms of term rewriting; see Sec. \ref{related-work})
%%If a type error is detected by an operator, the \emph{error denotation}, $\terr$, is returned instead of a valid denotation. In a practical implementation, a specialized error message and other diagnostic information would be provided when returning $\terr$, but we omit such details for simplicity. Terms of kind $\kDen$ can be \textsf{case} analyzed to determine if they are valid or errors, and if valid, to extract the translation and type.
%%
%%In the example in Fig. \ref{example}, the operator $\opvar{z}$ checks that no arguments were passed in using the simple helper function $\tvar{empty}$. If so, it returns a valid denotation by pairing the translation $\titerm{0}$ with the type $\ttype{Nat}{\tunit}$, as expected\footnote{Actually, there is no theoretical barrier to a different ``zero'' being used!}. If an argument was provided, the helper function returns $\terr$. The successor operator, $\opvar{s}$, takes one argument, so it pops a denotation off the argument list, making sure there are no more, and binds its translation and type to $\tvar{x}$ and $\tvar{t}$ respectively, all using the helper function $\tvar{pop\_final}$. It then checks that the argument's type is also $\ttype{Nat}{\tunit}$, returning a denotation pairing the translation $\titerm{\iup{\tvar{x}} + 1}$ with the type $\ttype{Nat}{\tunit}$ if so. The form $\iup{\tau}$ is used to ``un-reify'' reified internal terms, of kind $\kITerm$ (thus serving as the left-inverse of $\titerm{\iota}$). In this case, $\tvar{x}$ is a type-level variable of kind $\kITerm$ representing the translation of the argument to the successor operator, so we simply need to add one to it. Because we have checked that the denotation's type was $\ttype{Nat}{\tunit}$, and the representation schema will guarantee that expressions of this type always translate to integers, we know that it is safe to do so. If any of these steps fail, the various helper functions we use simply return $\terr$ (in practice, it would be prudent to equip each failure condition with a different error message). Compilation will also fail if we accidentally violate the representation schema, as we will see in Sec. \ref{repcon}.
%%
%%% The operator index, $\tauidx$, and . 
%%
%%\subsection{Functions, Variables and Arrow Types}
%%
%%The recursor operator, $\opvar{rec}$, proceeds similarly, extracting the translations and types of each of its three arguments. Of note, however, is how it handles the third argument, which binds two variables (the predecessor and the result of recursing on it). In \atlam, the built-in $\lambda$ operator serves as the sole mechanism for introducing bound variables. In most calculi, lambda terms have types of the form $\tau_1 \rightarrow \tau_2$. In \atlam, $\rightarrow$ corresponds to the built-in type family, $\fvar{Arrow}$. This family is indexed by a pair of types, $\kpair{\kTypeBlur}{\kTypeBlur}$, and is always in scope. Its representation schema simply maps to the corresponding internal arrow type, $\tvar{i}.\titype{\darrow{\trepof{\tfst{\tvar{i}}}}{\trepof{\tsnd{\tvar{i}}}}}$. The special form $\trepof{\tau}$ is used to refer, abstractly, to the representation of the type $\tau$ (we also see this used on line \ref{fix}). Although the $\lambda$ operator is built-in, because it needs to bind variables, application is just an operator associated with the $\fvar{Arrow}$ family, $\opvar{ap}$, as is seen on line \ref{ap}. The details are straightforward and given in the appendix.
%%
%%\subsection{Compilation}
%%\begin{figure}[t]
%%\small
%%\begin{mathpar}
%%\inferrule{
%%	\overbrace{\progOK{\emptyctx}{\fvalCtx_0}{\rho}}^{\text{\normalsize Kind Checking}}\\
%%	\overbrace{\pcompiles{\fvalCtx_0}{\rho}{\iota}}^{\text{\normalsize Active Typechecking and Translation}}
%%}{
%%	\ptcc{\rho}{\iota}
%%}
%%\end{mathpar}
%%\caption{\small Central Compilation Judgement of \atlam.}
%%\label{ccj}
%%\end{figure}
%%The \emph{central compilation judgement}, shown in Figure \ref{ccj}, captures the two phases of compilation: kind checking and active typechecking and translation. 
%%The first phase ensures that all type-level terms in the program are well-kinded and that all expressions and internal terms are closed. The kinding rules for programs are given in Figure \ref{kindprog}, and they rely on the kinding rules for type-level terms given in Figure \ref{tlkind}. These rules use contexts $\Sigma$ and $\Theta$ to track family and operator signatures, but beyond that, the rules are largely consistent with those of a simply-typed lambda calculus shifted into the type-level, with the addition of the handful of special forms constrained as described in the previous sections. The reader is encouraged to verify that the example in Fig. \ref{example} is well-kinded. 
%%% As we will see, the kind checking phase ensures that many kinds of errors are ruled out and that this process will not ``get stuck'' (see Sec. \ref{safety}). The evaluation semantics for type-level terms are given in Fig. \ref{tleval}.
%%\begin{figure}[t]
%%\small
%%$\fbox{\inferrule{}{\progOKX{\progsort}}}$
%%~~~$\tvarCtx ::= \emptyctx \pipe \tvarCtxX{t}{\kappa}$~~~
%%$\fCtx ::= \Sigma_0	 \pipe \fvalCtxX$
%%\begin{mathpar}
%%\inferrule[family-kinding]{
%%	\fvar{Fam} \notin \text{dom}(\fvalCtx)\\
%%	\kEq{\kappaidx}\\
%%	\opType{\tvarCtx}{\fvalCtxX}{\theta}{\Theta}\\\\
%%	\tKind{\tvarCtx, \tvar{i}:{\kappaidx}}{\fvalCtxX}{\tau}{\kIType}\\
%%	\progOK{\tvarCtx}{\fvalCtxX}{\rho}
%%}{
%%	\progOKX{\pfam{\familyDf}{\rho}}
%%}
%%
%%\inferrule[def-kinding]{
%%	\tKindX{\tau}{\kappa}\\
%%	\progOK{\tvarCtxX{t}{\kappa}}{\fvalCtx}{\rho}
%%}{
%%	\progOKX{\pdef{t}{\kappa}{\tau}{\rho}}
%%}
%%
%%\inferrule[exp-kinding]{
%%	\exprOK{\tvarCtx}{\emptyctx}{\fvalCtx}{e}
%%}{
%%	\progOKX{e}
%%}
%%\end{mathpar}
%%$\fbox{$\tKindX{\theta}{\Theta}$}$
%%~~~$\Theta ::= \kOpS{op}{\kappaidx} \pipe \Theta, \Theta$~~~
%%\begin{mathpar}
%%\inferrule[op-kinding]{
%%	\tKind{\tvarCtx, \tOfKind{\tvar{i}}{\kappai}, \tOfKind{\tvar{a}}{\klist{\kDen}}}{\fvalCtx}{\tau}{\kDen}
%%}{
%%	\opType{\tvarCtx}{\fvalCtx}{{\tops{op}{\kappaidx}{i}{a}{\tau}}}{
%%	\kOpS{op}{\kappaidx}}
%%}
%%
%%\inferrule[ops-kinding]{
%%	\tKindX{\theta_1}{\Theta_1}\\
%%	\tKindX{\theta_2}{\Theta_2}\\\\
%%	\text{dom}(\theta_1) \cap \text{dom}(\theta_2) = \emptyset
%%}{
%%	\tKindX{\theta_1; \theta_2}{\Theta_1, \Theta_2}
%%}
%%\end{mathpar}
%%$\fbox{$\exprOKX{e}$}$
%%~~~$\itvarCtx ::= \emptyctx \pipe \itvarCtx, \evar{x}$
%%\begin{mathpar}
%%\inferrule[e-var-kinding]{ }{
%%	\exprOK{\tvarCtx}{\itvarCtx, \evar{x}}{\fvalCtx}{\evar{x}}
%%}
%%
%%\inferrule[e-lam-kinding]{
%%	\tKindX{\tau}{\kTypeBlur}\\
%%	\exprOK{\tvarCtx}{\eivarCtxX{x}}{\fvalCtx}{e}
%%}{
%%	\exprOKX{\elam{\evar{x}}{\tau}{e}}
%%}
%%
%%\inferrule[e-op-kinding]{
%%	\fvarOfType{Fam}{\kappaidx}{\Theta} \in \fvalCtx\\
%%	\kOpS{op}{\kappai} \in \Theta\\
%%	\tKindX{\taui}{\kappai}\\\\
%%	\exprOKX{e_1}\\
%%	\cdots\\
%%	\exprOKX{e_n}
%%}{
%%	\exprOKX{\eop{Fam}{op}{\taui}{\splat{e}{1}{n}}}
%%}
%%\end{mathpar}
%%\caption{\small Kinding for programs. Variable contexts $\tvarCtx$ and $\itvarCtx$ obey standard structural properties. Kinding rules for type-level terms are given in Figure \ref{tlkind}.}
%%\label{kindprog}
%%\vspace{-10pt}
%%\end{figure}
%%%\subsection{Active Typechecking and Translation}
%%The second phase of compilation involves invoking the  logic implemented by user-defined operator families to typecheck and translate the program. The rules for this phase are given in Fig. \ref{att}. The context $\fvalCtx$ tracks the operator definitions and representation schemas associated with families in scope. Because the logic inside operators must be invoked during this phase, we need an evaluation semantics for type-level terms, given in Fig. \ref{tleval}. This is again a largely unsurprising collection of rules with the exception of \textsc{dencase-eval-valid}, which we will explain below.
%%
%%\subsection{Abstract Representations and Abstract Internal Types}\label{repcon}
%%To typecheck and translate an expression, $e$, the rule \textsc{att-exp} first assigns a type, $\tau$, and an \emph{abstract translation}, $\gabs$, to it, as determined by the judgement $\ecompilesX{e}{\tau}{\gabs}$. It then \emph{deabstracts} this abstract translation to complete the compilation process, as determined by the judgement $\eraseX{\gabs}{\iota}$. The purpose of the abstract translation phase is to ensure that the implementation details of type families are not exposed to other families, so that invariants that they rely on are preserved when families are composed. For example, our implementation of natural numbers as integers in Fig. \ref{example} maintains the invariant that the translation is non-negative. If the knowledge that natural numbers are implemented as integers was externally visible, a different extension could introduce an operator like $\tops{badnat}{\kunit}{i}{a}{\tapp{\tvar{const}}{\tapp{\tvar{a}}{\tden{\titerm{-1}}{\ttype{Nat}{\tunit}}}}}$, breaking this invariant (and thus the guarantee that the recursor always terminates!) 
%%%To put it another way, if a collection of operators can be shown to be a full and faithful (that is, adequate) encoding of a particular type system, then it does not matter what other types there are in the system.
%%
%%We preclude such operations by a mechanism similar to the abstract type mechanism supported by ML-style module systems \cite{pfpl}. By keeping a type family's representation schema private to the operators associated with it, they maintain full control over representation invariants. So, because it cannot be shown given the knowledge available in the type family containing $\opvar{badnum}$ that the internal type associated with $\ttype{Nat}{\tunit}$ is $\dint$, the translation $-1$ will not be permitted according to the rules we will give below. The only way to produce a term of type $\ttype{Nat}{\tunit}$ as the result of applying an operator not associated with the family $\fvar{Nat}$ is if that operator extracts the term from an argument (e.g. when projecting it out of an $n$-tuple), in which case the necessary invariants are inductively  maintained (see Sec. \ref{safety}). Translations extracted from denotations via \textsf{case} analysis are tracked during this phase as \emph{abstract internal terms}, $\tvalof{\tau_1}{\tau_2}$ (see rule \textsc{dencase-eval-valid}).%
%%
%%Let us first review how expressions are assigned types and abstract translations. Variables (\textsc{att-var}) translate directly to variables in the internal language, with the type determined by the typing context, $\iota$. Lambda terms (\textsc{att-lam}) also translate to lambda terms in the internal language and are given the appropriate $\fvar{Arrow}$ type by extending the typing context and proceeding recursively as is usual. The internal type of the argument, however, is left as $\trepof{\tau_1'}$, which is called the \emph{abstract representation} of $\tau_1'$. The actual representation of this type is only available from within operators in its family.
%%
%%Now we will consider the important operator application rule (\textsc{att-op}). This rule operates by extracting the appropriate operator definition, evaluating the operator index to a value, $\tauidx'$, and recursively assigning a type and abstract translation to each argument. From these, it constructs a list of denotations  and passes this list, along with the fully evaluated operator index, to the  operator implementation, $\tauop$. If this results in a valid denotation, compilation can proceed, but if an $\terr$ is produced, compilation will stop because no other rule will apply (in practice, we would display a type error at this point).
%%
%%Before a valid denotation produced in this way can be used, however, we check that it is \emph{representationally consistent}, meaning that the abstract translation, $\gabs$, is of an abstract internal type consistent with the abstraction representation of the type it is paired with, $\tau$. The premise $\ddbar{\Xi_0,\fvar{Fam}}{\fvalCtx}{\trepof{\tau}}{\sabs}$ determines the abstract internal type and the premise $\checkRC{\gtCtx}{\Xi_0,\fvar{Fam}}{\fvalCtx}{\gabs}{\sabs}$ performs the internal type checking. To do so, the variable context, $\iota$, which associates variables with types, must be converted to a context associating those variables with their corresponding abstract internal types, $\gtCtx$. The relevant judgements are defined in Fig. \ref{ait}. The schema context $\Xi$ is used to track which representation schemas are visible. In our calculus, $\Xi_0=\fvar{Arrow}$ is always visible alongside the schema of the family associated with the operator being considered. This could be exanded to support module-scoped visibility (we do not include modules in our calculus for simplicity), mutually-defined type families or type families that intentionally expose their representation schemas publicly (as $\fvar{Arrow}$ does).
%%
%%To clarify this fundamental mechanism let us examine how the expression $\eop{Nat}{s}{\tunit}{\eop{Nat}{z}{\tunit}{ }}$ will be processed during this phase. The inner application of $\opvar{z}$ will produce the abstract translation $0$ paired with the type $\ttype{Nat}{\tunit}$. Because the representation schema for $\fvar{Nat}$ is available, we have that $\ddbar{\Xi_0,\fvar{Nat}}{\fvalCtx}{\trepof{\ttype{Nat}{\tunit}}}{\dint}$ by \textsc{show-rep}, as needed. In contrast, if $\opvar{badnat}$ were used inside, the representation schema of $\fvar{Nat}$ would not be available, so we can only apply \textsc{hide-rep}, which does not give us enough information to give $-1$ a type.  
%%
%%When the compiler next considers the outer application of $\opvar{s}$, it will pass the denotation $\tden{\titerm{0}}{\ttype{Nat}{\tunit}}$ into its definition.  There, it binds the translation to the variable $\tvar{x}$ (within the $\tvar{pop\_final}$ helper function). However, $\tvar{x}$ is not simply $\titerm{0}$. Instead, its provenance is tracked by using the form $\tvalof{\titerm{0}}{\ttype{Nat}{\tunit}}$. When attempting to check that $\iup{\tvar{x}} + 1$ is valid, the \textsc{show-trans} rule will reveal that it is an integer because, again, the appropriate representation schema is available. If the schema were not available, the most that could be derived is that $\checkRC{\gtCtx}{\Xi}{\fvalCtx}{\iup{\tvar{x}}}{\trepof{\ttype{Nat}{\tunit}}}$. The fact that natural numbers are represented using integers is not derivable (though it is true), so the addition operation would fail to type. This fact would, however, be sufficient for implementing families, like $\fvar{Ntuple}$, that do not require knowledge about a type's representation. We encourage the reader to derive these judgements for the logic in the $\opvar{rec}$ operator to strengthen their understanding of this fundamental mechanism.
%%
%%We cannot leave abstract representations and abstract internal terms in the result of compilation, so a final deabstraction phase erases these, replacing them with their underlying internal types and terms in all contexts. Just as with abstract types in modules or existential types, there is no run-time overhead to this mechanism -- programs run at full speed. We give the deabstraction rules in the appendix due to their simplicity.
%%
%
%%
%
%%
%%\begin{figure}[t]
%%\small
%%$\fbox{\inferrule{}{\concrep{\fvalCtx}{\tau}{\sigma}}}$
%%\begin{mathpar}
%%\inferrule[get-rep]{
%%	\fval{Fam}{\theta}{i}{\tau} \in \fvalCtx\\
%%	\tEvalX{[\tauidx/\tvar{i}]\tau}{\titype{\sigma}}\\
%%	\ddbarX{\sigma}{\sconc}
%%}{
%%	\concrep{\fvalCtx}{\ttype{Fam}{\tauidx}}{\sconc}
%%}
%%\end{mathpar}
%%$\fbox{\inferrule{}{\ddbarX{\sigma}{\sigma'}}}$
%%~~~$\fvalCtx ::= \fvalCtx_0 \pipe \fvalCtx, \fvalDf$~~~
%%~~~$\Xi ::= \Xi_0 \pipe \Xi, \fvar{Fam}$~~~
%%~~~$\Xi_0 := \fvar{Arrow}$
%%\begin{mathpar}
%%\inferrule[abs-int]{ }{
%%	\ddbarX{\dint}{\dint}
%%}
%%
%%\inferrule[abs-arrow]{
%%	\ddbarX{\sigma_1}{\sigma_1'}\\
%%	\ddbarX{\sigma_2}{\sigma_2'}
%%}{
%%	\ddbarX{\darrow{\sigma_1}{\sigma_2}}{\darrow{\sigma_1'}{\sigma_2'}}
%%}
%%
%%\inferrule[abs-prod]{
%%	\ddbarX{\sigma_1}{\sigma_1'}\\
%%	\ddbarX{\sigma_2}{\sigma_2'}
%%}{
%%	\ddbarX{\dpair{\sigma_1}{\sigma_2}}{\dpair{\sigma_1'}{\sigma_2'}}
%%}
%%
%%\inferrule[itype-inverse]{
%%	\ddbarX{\sigma}{\sabs}
%%}{
%%	\ddbarX{\dup{\titype{\sigma}}}{\sabs}
%%}
%%
%%\inferrule[show-rep]{
%%	\fvar{Fam} \in \Xi\\
%%	\concrep{\fvalCtx}{\ttype{Fam}{\tauidx}}{\sconc}
%%}{
%%	\ddbarX{\trepof{\ttype{Fam}{\tauidx}}}{\sconc}
%%}
%%
%%\inferrule[hide-rep]{
%%	\fvar{Fam} \notin \Xi
%%}{
%%	\ddbarX{\trepof{\ttype{Fam}{\tauidx}}}{\trepof{\ttype{Fam}{\tauidx}}}
%%}
%%\end{mathpar}
%%$\fbox{\inferrule{}{\eCtxTogCtxX{\etCtx}{\gtCtx}}}$
%%~~~$\gtCtx ::= \emptyctx \pipe \gtCtxX{x}{\sigma}$
%%\begin{mathpar}
%%\inferrule[abs-empty]{ }{
%%	\eCtxTogCtxX{\emptyctx}{\emptyctx}
%%}
%%
%%\inferrule[abs-ctx]{
%%	\eCtxTogCtxX{\etCtx}{\gtCtx}\\
%%	\ddbarX{\trepof{\tau}}{\sigma}
%%}{
%%	\eCtxTogCtxX{\etCtxX{x}{\tau}}{\gtCtxX{x}{\sigma}}
%%}
%%\end{mathpar}
%%$\fbox{\inferrule{}{\checkRCX{\iota}{\sigma}}}$
%%\begin{mathpar}
%%\inferrule[abs-i-var]{ }{
%%	\checkRC{\gtCtxX{x}{\sigma}}{\Xi}{\fvalCtx}{\evar{x}}{\sigma}
%%}
%%
%%\inferrule[abs-i-lam]{
%%	\ddbarX{\sigma_1}{\sigma_1'}\\
%%	\checkRC{\gtCtxX{x}{\sigma_1'}}{\Xi}{\fvalCtx}{\iota}{\sigma_2}
%%}{
%%	\checkRCX{\ilam{x}{\sigma_1}{\iota}}{\darrow{\sigma_1'}{\sigma_2}}
%%}
%%
%%\inferrule[abs-i-ap]{
%%	\checkRCX{\iota_1}{\darrow{\sigma_1}{\sigma_2}}\\
%%	\checkRCX{\iota_2}{\sigma_1}
%%}{
%%	\checkRCX{\iapp{\iota_1}{\iota_2}}{\sigma_2}
%%}
%%
%%\inferrule[abs-i-fix]{
%%	\ddbarX{\sigma}{\sigma'}\\
%%	\checkRC{\gtCtxX{x}{\sigma'}}{\Xi}{\fvalCtx}{\iota}{\sigma'}
%%}{
%%	\checkRCX{\ifix{x}{\sigma}{\gabs}}{\sigma'}
%%}
%%\\
%%\text{\color{gray} (standard statics for integers and products omitted)}
%%\\
%%%\inferrule[abs int]{ }{
%%%	\checkRCX{\iintlit}{\dint}
%%%}
%%%
%%%\inferrule[abs op]{
%%%	\checkRCX{\iota_1}{\dint}\\
%%%	\checkRCX{\iota_2}{\dint}
%%%}{
%%%	\checkRCX{\iop{\iota_1}{\iota_2}}{\dint}
%%%}
%%%
%%%\inferrule[abs pair]{
%%%	\checkRCX{\iota_1}{\sigma_1}\\
%%%	\checkRCX{\iota_2}{\sigma_2}
%%%}{
%%%	\checkRCX{\ipair{\iota_1}{\iota_2}}{\dpair{\sigma_1}{\sigma_2}}
%%%}
%%%
%%%\inferrule[abs fst]{
%%%	\checkRCX{\iota}{\dpair{\sigma_1}{\sigma_2}}
%%%}{
%%%	\checkRCX{\ifst{\iota}}{\sigma_1}
%%%}
%%%
%%%\inferrule[abs snd]{
%%%	\checkRCX{\iota}{\dpair{\sigma_1}{\sigma_2}}
%%%}{
%%%	\checkRCX{\isnd{\iota}}{\sigma_2}
%%%}
%%%
%%% \inferrule[abs-int-eq]{
%%% 	\checkRCX{\iota_1}{\dint}\\
%%% 	\checkRCX{\iota_2}{\dint}\\\\
%%% 	\checkRCX{\iota_3}{\sigma}\\
%%% 	\checkRCX{\iota_4}{\sigma}
%%% }{
%%% 	\checkRCX{\iIfEq{\iota_1}{\iota_2}{\dint}{\iota_3}{\iota_4}}{\sigma}
%%% }
%%\inferrule[abs-iterm-inverse]{
%%	\checkRCX{\iota}{\sigma}
%%}{
%%	\checkRCX{\iup{\titerm{\iota}}}{\sigma}
%%}
%%
%%\inferrule[show-trans]{
%%	\fvar{Fam} \in \Xi\\
%%	\concrep{\fvalCtx}{\ttype{Fam}{\tauidx}}{\sigma}\\
%%	\checkRCX{\iota}{\sigma}
%%}{
%%	\checkRCX{\tvalof{\titerm{\iota}}{\ttype{Fam}{\tauidx}}}{\sigma}
%%}
%%
%%\inferrule[hide-trans]{
%%	\fvar{Fam} \notin \Xi\\
%%	\concrep{\fvalCtx}{\ttype{Fam}{\tauidx}}{\sigma}\\
%%	\checkRCX{\iota}{\sigma}
%%}{
%%	\checkRCX{\tvalof{\titerm{\iota}}{\ttype{Fam}{\tauidx}}}{\trepof{\ttype{Fam}{\tauidx}}}
%%}
%%\end{mathpar}
%%\caption{\small Abstracted internal typing}
%%\label{ait}
%%\end{figure}
%%%
%%%family RECORD of (string, type) dict {
%%%  new(idx, args. foldl (fn (x, y) => 
%%%  get[field : string](idx. case find(idx, field) of SOME t => t | _ => err)
%%
%%\section{Safety of @$\lambda$}\label{safety}
%%In giving users of a language direct influence on the typechecking and translation of expressions, it is essential to consider safety properties. It must not be possible for well-typed programs to fail to finish compiling or go wrong at run-time because of a buggy extension. It is also considered desirable for typechecking to be decidable. And for extensions to be reliable, they must not be allowed to interfere with one another under any circumstance. By carefully designing our extension mechanism, we believe we have achieved each of these goals.
%%
%%\subsection{Representational Consistency and Type Safety}
%%We do not need to give a semantics to internal terms and internal types that do not survive deabstraction. The remaining terms and types form a variant of PCF with well-understood primitives (here, integers and binary products) known to be type safe \cite{pfpl}. If compilation can be shown to always result in a well-typed term of this language, then type safety follows by composing these two facts. Fortunately, this fact is a corollary of our representational consistency mechanism, described in Sec. \ref{repcon}. The proof is by straightforward (though deeply nested!) rule induction.
%%
%%%\begin{theorem}[Representational Consistency]
%%%If $\progOK{ }{\Sigma}{e}$ and $\ecompiles{ }{\fvalCtx}{e}{\tau}{\gabs}$ and $\eraseX{\gabs}{\iota}$ and 	$\ddbar{\Xi_0}{\fvalCtx}{\trepof{\tau}}{\sabs}$ and $\eraseX{\sabs}{\sigma}$ then $\checkRC{ }{\Xi_0}{\fvalCtx}{\iota}{\sigma}$. 
%%%\end{theorem}
%%
%%\subsection{Decidability}
%%To show that compilation is decidable, we need to show that both kind checking and active typechecking and compilation are decidable. This hinges on showing that the type-level language is typesafe and that evaluation of type-level terms always terminates. 
%%
%%%\begin{theorem}[Kind Safety and Termination]
%%%If $\vdash_{\Sigma} \tau : \kappa$ then $\tEvalX{\tau}{\tau'}$ such that $\vdash_{\Sigma_0} \tau' : \kappa$ and $\tEvalX{\tau'}{\tau'}$. 
%%%\end{theorem}
%%
%%The proof is straightforward because the type-level language is based on a conventional simply-typed lambda calculus with only bounded recursion over lists, so standard techniques (e.g. logical relations) can be used directly. The only new constructs are the types, denotations and reified internal language forms, but because none of these can contain type-level functions by the statics, there is no risk of introducing self-reference by their inclusion. If the type-level language included constructs like inductive datatypes (without a strict positivity condition) or general recursion, this theorem would be weakened.
%%
%%\subsection{Non-Interference}
%%Our abstraction mechanism guarantees that the representation invariants collectively maintained by the operators associated with a type family cannot be violated by operators in other type families, by ensuring that introductory forms cannot be defined outside the family. One way to state this is as follows:
%%
%%%\begin{theorem}[Non-Interference]
%%%For any property $P(\iota)$, if $\progOK{ }{\Sigma}{e}$ and $\ecompiles{ }{\fvalCtx}{e}{\tau}{\gabs}$ and $\eraseX{\gabs}{\iota}$ implies $P(\iota)$, then for any $\Sigma'$ and $\fvalCtx'$ disjoint from $\Sigma$ and $\fvalCtx$, we have that if $\progOK{ }{\Sigma,\Sigma'}{e'}$ and $\ecompiles{ }{\fvalCtx,\fvalCtx'}{e'}{\tau}{\gabs'}$ and $\eraseX{\gabs'}{\iota'}$ then $P(\iota')$.
%%%\end{theorem}
%%The proof relies on the fact that the show/hide rules are mutually exclusive and so there are two cases to consider for each form of expression, and weakening properties of the family contexts. This powerful theorem allows us to compose type families arbitrarily without needing to handle cases in our implementation that are ruled out based on a local analysis. The details will be provided in the appendix.
%%
%%% \subsection{Modular Verification}
%%% nope
%%
%%% that's all folks
%%
%%\section{Use Cases}\label{examples}
%%
%%Aldrich argues that dynamic dispatch has proven useful for building  large software frameworks \cite{aldrich-onward13}. Object-oriented languages support dynamic dispatch directly but encodings of dynamic dispatch in terms of standard functional primitives can be awkward and difficult to reason about (e.g. \cite{pfpl} Ch. X).
%%
%%Indexed type families and indexed operator families are a powerful tool for programming language semantics. In \emph{Practical Foundations for Programming Languages}, for example, nearly every chapter defines a new collection of indexed types and operators and gives their semantics in isolation from the constructs in the other chapters \cite{pfpl}. For example, $n$-ary sums and products are families indexed by a list of types. Labeled variants of these simply pair the labels with the types as static strings. Even nested pattern matching, such as that found in modern functional languages, can be understood as an operator family indexed by a series of patterns, which can be represented as type-level data. In the appendix, we show how to implement sums and products in \atlam. 
%%
%%The book also shows constructs as varied as dynamic types as well as a number of constructs for parallelism and concurrency, such as futures, promises and actors. With a sufficiently capable internal language (exposing basic concurrency primitives, for example), each of these could be fully implemented as libraries using our mechanism. Few languages include primitive support for several such abstractions, even though they are often useful together. In a preliminary implementation of this calculus, Ace, which is beyond the scope of this paper, we have implemented the entirety of the OpenCL programming language as a library, along with primitives supporting partitioned global address spaces.
%%
%%Object systems too could be implemented using this mechanism, with indices serving to capture the inheritance data and the signatures. Operator families for reading and writing fields and/or sending messages would be parameterized by strings naming them. The operator definition would simply search the signatures going up the inheritance hierarchy, and implement objects in a conventional way (e.g. using a v-table together with a pointer to a structure containing field data). A variety of object systems could coexist within the same language. Another related example would be to implement a safe and efficient interoperability layer with an existing OO language by capturing its type system as an extension, to be used only at language boundaries.
%%
%%Finally, a variety of domain-specific type systems, capturing complex rule systems like the system of scientific units of measure (a family indexed by the measure being used and the type which the measure is being applied to), regular expressions \cite{regexp} (a family indexed by the number of captured groups) or XML \cite{xml} (indexed by a document schema) would be possible. Recent work examined a specialized type system capturing the semantics of the widely-used jQuery javascript library \cite{jquery-oopsla}. Today, these all require \emph{ad hoc} language-external solutions, or the development of new languages. 
%%
%%%\section{Design Considerations}\label{design}
%%
%\section{Related Work}\label{related-work}
%Our representation schema abstraction mechanism relates closely to abstract and existential types \cite{pfpl,atpl}. Our calculus enforces the abstraction barriers  in a purely syntactic manner, as in previous work on syntactic type abstraction \cite{syntypeabs}. While this work is all focused on abstracting away the identity of a particular type outside of the ``principal'' it is associated with (e.g. a module), we focus on abstracting away the knowledge of how a primitive type family is implemented outside of a limited scope.
%
%The representational consistency mechanism brings into the language work on typed compilation, especially work done on Standard ML in the TILT project \cite{tilt}. Indeed, the specification of Standard ML is structured around a typed internal language and a judgement that assigns a type and an internal term to each expression \cite{smlstd}. Representational consistency is related to the notion of type-directed copmilation in this work.
%
%Type-level computation is supported in some form by a growing number of languages. For example, Haskell supports a simple form of it \cite{Chakravarty:2005:ATC}). Ur uses type-level records and names to support typesafe metaprogramming, with applications to web programming \cite{conf/pldi/Chlipala10}. $\Omega$mega adds algebraic data types at the type-level, using these to increase the expressive power of algebraic data types at the expression level \cite{conf/cefp/SheardL07}. Dependently-typed languages blur the traditional phase separation between types and expressions, so type-level computation is often implicitly used (though not always in its most general form, e.g. Deputy \cite{conf/icfp/ChenX05}, ATS \cite{conf/esop/ConditHAGN07}). We show how to integrate language extensions into the type-level language, drawing on ideas about \cite{activelibraries}.
%
%% \subsection{Run-Time Indirection}
%% {\it Operator overloading} \cite{vanWijngaarden:Mailloux:Peck:Koster:Sintzoff:Lindsey:Meertens:Fisker:acta:1975} and {\it metaobject dispatch} \cite{Kiczales91} are run-time protocols that translate operator invocations into function calls. The function is typically selected according to the type or value of one or more operands. These protocols share the notion of {\it inversion of control} with type-level specification. However, type-level specification is a {\it compile-time} protocol focused on enabling specialized verification and implementation strategies, rather than simply enabling run-time indirection.
%
%Many languages and tools allow developers to rewrite expressions according to custom rules. These can broadly be classified as {\it term rewriting systems}. Macro systems, such as those characteristic of the LISP family of languages \cite{mccarthy1978history}, are the most prominent example. Some compile-time metaprogramming systems also allow users to manipulate syntax trees (e.g. MetaML \cite{Sheard:1999:UMS}), and external rewrite systems also exist for many languages.
%These facilities differ from AT\&T in that they involve direct manipulation of terms, while AT\&T involves extending typechecking and translation logic directly. We also draw a distinction between the type-level, used to specify types and compile-time logic, the expression grammar, used to describe run-time behavior, and the internal language, used to implement this behavior. By doing so, each component can be structured and constrained as appropriate for its distinct role, as we have shown.
%
%Previous work on extensible languages has suffered from problems with either expressiveness or safety. For example, a number of projects, such as SugarJ \cite{erdweg2011sugarj}, allow for user-defined desugarings (and indeed, our system would clearly benefit from integration with such a mechanism), but this does not allow the semantics to be fundamentally extended nor for implementation details to be hidden. Recent variants of this work has investigated introducing new typing rules, but only if they are admissible by the base type system \cite{erdwegicfp}. Our work allows for entirely new logic to be added to the system, requiring only that the implementation of this logic respect the internal type system. A number of other extensible languages (e.g. Xroma \cite{xroma}) and compilers do allow new static semantics to be introduced, but in an unconstrained manner based on global pattern matching and rewriting, leading to significant problems with interference and safety. Our work aims to provide a sound and safe underpinning, based around well-understood concepts of type and operator families, to language extensibility.
%
%In the future, we will investigate mechanisms to enable type systems with specialized binding and scoping rules, as well as integration of dependent kinds to support mechanized verification of key properties like representational consistency and adequacy against a declarative specification at kind-checking time. We will also investigate mechanisms that enable a more natural syntax (e.g. Wyvern's type-directed syntax \cite{globaldsl}). 
%
%%\item With type-level specification, dispatch to a type-level function occurs implicitly on the basis of the structure of an expression. In contrast, most term-rewriting systems operate by  explicit invocation of a macro or specialized syntax. Some LISP macro systems have explored pattern-based dispatch (e.g. A*\cite{fowler2010domain}, EPP\cite{fowler2010domain}) and macro systems for object-oriented languages, like OpenC++ \cite{fowler2010domain} and OpenJava \cite{fowler2010domain}, do offer a somewhat limited form of operation-based dispatch.
%
%% \subsection{Language Frameworks}
%% When the mechanisms available in an existing language prove insufficient, researchers and domain experts must design a new language. A number of tools have been developed to assist with this task, including compiler generators, language workbenches and domain-specific language frameworks (cf \cite{fowler2010domain}).
%
%% A major barrier to adoption is the fact that interoperability is intrinsically problematic. Even languages which target a common platform, such as the Java Virtual Machine, can only interact using its limited set of primitives. Specialized typing rules are not checked at language boundaries, performance often suffers, and the syntax can be unnatural, particularly for languages which differ significantly from the platform's native language (e.g. Java).
%
%% Instead of focusing on defining standalone languages, type-level specification gives greater responsibility in a granular manner to libraries. In this way, a range of constructs can coexist within the same program and, assuming that it can be shown by some method that various constructs are safely composable, be mixed and matched. The main limitation is that the protocol requires defining a fixed source grammar, whereas a specialized language has considerable flexibility in that regard. Nevertheless, as Ace shows, a simple grammar can be used quite flexibly.
%% \subsection{Extensible Compilers}
%% An alternative methodology is to implement language features granularly as compiler extensions. As discussed in Section 1, existing designs suffer from the same problems related to composability, modularity\-, safety and security as extensible languages, while also adding the issue of language fragmentation.
%
%% Type-level specification can in fact be implemented within a compiler, rather than provided as a core language feature. This would resolve some of the issues, as described in this paper. However, by leveraging type-level computation to integrate the protocol directly into the language, we benefit from common module systems and other shared infrastructure. We also avoid the fragmentation issue.
%% \subsection{Specification Languages}
%% Several {\it specification languages} (or {\it logical frameworks}) based on these theoretical formulations exist, including the OBJ family of languages (e.g. CafeOBJ \cite{Diaconescu-Futatsugi01}). They provide support for verifying a program against a language specification, and can automatically execute these programs as well in some cases. The  language itself specifies which verification and execution strategies are used.
%
%% Type-determined compilation takes a more concrete approach to the problem, focusing on combining {\it implementations} of different\- logics, rather than simply their specifications. In other words, it focuses on combining {\it type checkers} and {\it implementation strategies} rather than more abstract representations of a language's type system and dynamic semantics. In Section 4, we outlined a preliminary approach based on proof assistant available for the type-level language to unify these approaches, and we hope to continue this line of research in future work.
%
%% CANT GUARANTEE THAT SPECIFICATIONS ARE ACTUALLY DECIDABLE
%\begin{figure}
%\small
%$\fbox{$\tEvalX{\tau}{\tau'}$}$
%\begin{mathpar}
%\inferrule[n-t-lam]{ }{
%	\tEvalX{\tlam{t}{\kappa}{\tau}}{\tlam{t}{\kappa}{\tau}}
%}
%
%\inferrule[n-t-ap]{
%	\tEvalX{\tau_1}{\tlam{t}{\kappa}{\tau}}\\
%	\tEvalX{\tau_2}{\tau_2'}\\
%	\tEvalX{\subst{\tau_2'}{\tvar{t}}{\tau}}{\tau'}
%}{
%	\tEvalX{\tapp{\tau_1}{\tau_2}}{\tau'}
%}
%
%\text{\color{gray} (normalization for integers, labels, lists, products and sums also standard)}
%
%\inferrule[n-t-eq-equal]{
%	\tEvalX{\tau_1}{\tau_1'}\\
%	\tEvalX{\tau_2}{\tau_1'}\\
%	\tEvalX{\tau_3}{\tau_3'}
%}{
%	\tEvalX{\tifeq{\tau_1}{\tau_2}{\kappa}{\tau_3}{\tau_4}}{\tau_3'}
%}
%
%\inferrule[n-t-eq-disequal]{
%	\tEvalX{\tau_1}{\tau_1'}\\
%	\tEvalX{\tau_2}{\tau_2'}\\\\
%	\tau_1' \neq \tau_2'\\
%	\tEvalX{\tau_4}{\tau_4'}
%}{
%	\tEvalX{\tifeq{\tau_1}{\tau_2}{\kappa}{\tau_3}{\tau_4}}{\tau_4'}
%}
%
%\inferrule[nt-ty]{
%	\tEvalX{\tauidx}{\tauidx'}
%}{
%	\tEvalX{\ttype{Tycon}{\tauidx}}{\ttype{Tycon}{\tauidx'}}
%}
%
%\inferrule[nt-tycase-match]{
%	\tEvalX{\tau}{\ttype{Tycon}{\tauidx}}\\
%	\tEvalX{\subst{\tauidx}{\tvar{x}}{\tau_1}}{\tau_1'}
%}{
%	\tEvalX{\tfamcase{\tau}{Tycon}{x}{\tau_1}{\tau_2}}{\tau_1'}
%}
%
%\inferrule[nt-tycase-fail]{
%	\tEvalX{\tau}{\ttype{Tycon'}{\tauidx}}\\
%	\fvar{Tycon} \neq \fvar{Tycon'}\\
%	\tEvalX{\tau_2}{\tau_2'}
%}{
%	\tEvalX{\tfamcase{\tau}{Tycon}{x}{\tau_1}{\tau_2}}{\tau_2'}
%}
%
%\inferrule[nt-d]{
%	\tEvalX{\tau}{\tau'}\\
%	\tiEvalX{\ibar}{\ibar'}
%}{
%	\tEvalX{\tden{\ibar}{\tau}}{\tden{\ibar'}{\tau'}}
%}
%
%\inferrule[nt-d-checked]{
%	\tEvalX{\tau}{\tau'}\\
%	\tiEvalX{\ibar}{\ibar'}
%}{
%	\tEvalX{\tden{\ibar}{\tau}^\checkmark}{\tden{\ibar'}{\tau'}^\checkmark}
%}
%
%\inferrule[nt-typeof]{
%	\tEvalX{\tau}{\tden{\ibar}{\tau'}}
%}{
%	\tEvalX{\ttypeof{\tau}}{\tau'}
%}
%
%\inferrule[nt-typeof-checked]{
%	\tEvalX{\tau}{\tden{\ibar}{\tau'}^\checkmark}
%}{
%	\tEvalX{\ttypeof{\tau}}{\tau'}
%}
%
%\inferrule[nt-reptype]{
%	\tiEvalX{\sbar}{\sbar'}
%}{
%	\tEvalX{\titype{\sbar}}{\titype{\sbar'}}
%}
%\end{mathpar}
%$\fbox{$\tiEvalX{\ibar}{\ibar'}$}$
%\begin{mathpar}
%\inferrule[nt-i-var]{ }{
%	\tiEvalX{\evar{x}}{\evar{x}}
%}
%~~~~~
%%\inferrule[nt-i-lam]{
%%	\tiEvalX{\sigma}{\sigma'}\\
%%	\tiEvalX{\iota}{\iota'}
%%}{
%%	\tiEvalX{\ilam{\evar{x}}{\sigma}{\iota}}{\ilam{\evar{x}}{\sigma'}{\iota'}}
%%}
%%
%\inferrule[nt-i-fix]{
%	\tiEvalX{\sbar}{\sbar'}\\
%	\tiEvalX{\ibar}{\ibar'}
%}{
%	\tiEvalX{\ifix{\evar{x}}{\sbar}{\ibar}}{\ifix{\evar{x}}{\sbar'}{\ibar'}}
%}
%~~~~~
%\inferrule[nt-i-transof]{
%	\tEvalX{\tau}{\tau'}
%}{
%	\tiEvalX{\itransof{\tau}}{\itransof{\tau'}}
%}
%\\
%\text{\color{gray} (omitted forms have similarly recursive rules)}
%%\inferrule[iterm unquote eval]{
%%	\tEvalX{\tau}{\tau'}
%%}{
%%	\tiEvalX{\iup{\tau}}{\iup{\tau'}}
%%}
%%
%\end{mathpar}
%$\fbox{$\tiEvalX{\sbar}{\sbar'}$}$
%\begin{mathpar}
%%\inferrule[i-int-eval]{ }{
%%	\tiEvalX{\dint}{\dint}
%%}
%%
%%\inferrule[i-prod-eval]{
%%	\tiEvalX{\sigma_1}{\sigma_1'}\\
%%	\tiEvalX{\sigma_2}{\sigma_2'}
%%}{
%%	\tiEvalX{\dpair{\sigma_1}{\sigma_2}}{\dpair{\sigma_1'}{\sigma_2'}}
%%}
%%
%\inferrule[nt-i-parr]{
%	\tiEvalX{\sbar_1}{\sbar_1'}\\
%	\tiEvalX{\sbar_2}{\sbar_2'}
%}{
%	\tiEvalX{\darrow{\sbar_1}{\sbar_2}}{\darrow{\sbar_1'}{\sbar_2'}}
%}
%
%\inferrule[nt-i-unquote]{
%	\tEvalX{\tau}{\tau'}
%}{
%	\tiEvalX{\dup{\tau}}{\dup{\tau'}}
%}
%
%\inferrule[nt-i-repof]{
%	\tEvalX{\tau}{\tau'}
%}{
%	\tiEvalX{\trepof{\tau}}{\trepof{\tau'}}
%}
%
%\text{\color{gray} (omitted forms have similarly recursive rules)}
%\end{mathpar}
%\caption{\small Normalization semantics for type-level terms, including those inside translational IL. There is nothing clever happening here.}
%\label{tleval}
%\end{figure}

\bibliographystyle{abbrv}
\bibliography{../research}

\begin{figure*}[p!]
\small
{\color{gray}\text{(constructs lifted from $\mathcal{L}\{{\rightarrow}\,{\forall}\,{\mu_\text{ind}}\,{\kunit}\,{\times}\,{+}\}$ are standard, omitted \cite{pfpl})}}~\hfill\fbox{$\sofk{\kDelta}{\kGamma}{\kXi}{\st}{\kappa\moutput}$}
~\fbox{$\sqofkX{\qcolor{\ity}}$}~\fbox{$\sqofkX{\qcolor{\itm}}$}
\begin{mathpar}
\inferrule[k-ty]{
  c[\ktyidx] \in \kXi\\\\
  \sofkX{\sttyidx}{\ktyidx}
}{
  \sofkX{\sty{c}{\sttyidx}}{\kty}
}

\inferrule[k-otherty]{ }{
  \sofkX{\sotherty{m}{\tau}}{\kty}
}

\inferrule[k-tycase]{
  c[\ktyidx] \in \kXi\\\\
  \sofkX{\st}{\kty}\\
  \sofk{\kDelta}{\kGamma, \sx : \ktyidx}{\kXi}{\st_1}{\kappa}\\
  \sofkX{\st_2}{\kappa}
}{
  \sofkX{\stycase{c}{\st}{\sx}{\st_1}{\st_2}}{\kappa}
}

\inferrule[k-qity]{\sqofkX{\qcolor{\ity}}}{
  \sofkX{\sqity{\qcolor{\ity}}}{\kity}
}

\inferrule[k-qity-uq]{
  \sofkX{\st}{\kity}
}{
  \sqofkX{\qcolorc\qtuq{\color{black}\st\qcolorc}\color{black}}
}

\inferrule[k-rep]{
  \sofkX{\st}{\kty}
}{
  \sofkX{\srep{\st}}{\kity}
}

%\inferrule[kity]{a}{b}
%
\inferrule[k-qitm]{
  \sqofkX{\qcolor{\itm}}
}{
  \sofkX{\sqitm{\qcolor{\itm}}}{\kitm}
}

\inferrule[k-qitm-uq]{
  \sofkX{\st}{\kitm}
}{
  \sqofkX{\qcolorc\quq{\color{black}\st\qcolorc}\color{black}}
}

\text{\color{gray} (rules for remaining \qcolor{quoted forms} generically recursive)}

\inferrule[k-syn]{
  n' < n
}{
  \sofkX{\ssyn{n'}}{\kty \times \kitm}
}

%\inferrule[kitm]{a}{b}
%
\inferrule[k-ana]{
  n' < n\\
  \sofkX{\st}{\kty}
}{
  \sofkX{\sana{n'}{\st}}{\kitm}
}

\inferrule[k-raise]{
  \kok{\kDelta}{\emptyset}{\kappa}
}{
  \sofkX{\sraise{\kappa}}{\kappa}
}

\end{mathpar}
\caption{Kinding for static terms.}
\label{statics-SL}
\end{figure*}
\begin{figure*}[p!]
\small
$\begin{array}{llll}
\mvlbl{concrete rep whitelist} & \mvlbl{abstract rep store} & \mvlbl{argument store} & \mvlbl{operation context}\\
\boxdot ::= {\rightharpoonup} ~|~ \boxdot, \tc \ & \memD ::= \emptyset ~|~ \memD, \st \leftrightsquigarrow \alpha & \memG ::= \cdot ~|~ \memG, n \hookrightarrow e ~|~ \memG, n \hookrightarrow e : \st \leadsto \iota/x : \tau & \aCtx ::= \boxdot; \Upsilon; \Phi 
\end{array}$~\hfill{\fbox{$\snorm{\st}{\memD}{\memG}{\aCtx}{\st\moutput}{\memD\moutput}{\memG\moutput}$}}\\
{\color{gray} (constructs lifted from $\mathcal{L}\{{\rightarrow}\,{\forall}\,{\mu_\text{ind}}\,{\kunit}\,{\times}\,{+}\}$ standard, stores propagate as suggested by rules below, error propagation rules omitted \cite{pfpl})}
~\hfill\fbox{$\serrX{\st}$}
\begin{mathpar}
\inferrule[n-ty]{
  \snormX{\st}{\st'}
}{
  \snormX{\sty{c}{\st}}{\sty{c}{\st'}}
}

\inferrule[n-otherty]{ }{
  \snorm{\sotherty{m}{\tau}}{\memD}{\memG}{\aCtx}{\sotherty{m}{\tau}}{\memD}{\memG}
}

\inferrule[n-tycase-1]{
  \snorm{\st}{\memD}{\memG}{\aCtx}{\sty{c}{\sttyidx}}{\memD'}{\memG'}\\\\
  \snorm{[\sttyidx/\svar{x}]\st_1}{\memD'}{\memG'}{\aCtx}{\st_1'}{\memD''}{\memG''}
}{
  \snorm{\stycase{c}{\st}{\svar{x}}{\st_1}{\st_2}}{\memD}{\memG}{\aCtx}{\st_1'}{\memD''}{\memG''}
}

\inferrule[n-tycase-2]{
  \snorm{\st}{\memD}{\memG}{\aCtx}{\st'}{\memD'}{\memG'}\\\\
  \st' \neq \sty{c}{\sttyidx}\\
  \snorm{\st_2}{\memD'}{\memG'}{\aCtx}{\st_2'}{\memD''}{\memG''}
}{
  \snorm{\stycase{c}{\st}{\svar{x}}{\st_1}{\st_2}}{\memD}{\memG}{\aCtx}{\st_2'}{\memD''}{\memG''}
}

\inferrule[n-qity-parr]{
  \snorm{\sqity{\qcolor{\tau_1}}}{\memD}{\memG}{\aCtx}{\sqity{\tau_1'}}{\memD'}{\memG'}\\\\
  \snorm{\sqity{\qcolor{\tau_2}}}{\memD'}{\memG'}{\aCtx}{\sqity{\tau_2'}}{\memD''}{\memG''}
}{
  \snorm{\sqity{\qcolor{\tau_1 \rightharpoonup \tau_2}}}{\memD}{\memG}{\aCtx}{\sqity{\tau_1' \rightharpoonup \tau_2'}}{\memD''}{\memG''}
}

\inferrule[n-qity-alpha]{ }{
  \snorm{\sqity{\qcolor{\alpha}}}{\memD}{\memG}{\aCtx}{\sqity{\alpha}}{\memD}{\memG}
}

\inferrule[n-qity-forall]{
  \snormX{\sqity{\qcolor{\tau}}}{\sqity{\tau'}}
}{
  \snormX{\sqity{\qcolor{\iforall{\alpha}{\tau}}}}{\sqity{\iforall{\alpha}{\tau'}}}
}

\text{\color{gray}(other \qcolor{quoted internal types} analagously recursive)}

\inferrule[n-qity-uq]{
  \snormX{\st}{\sqity{\tau}}
}{
  \snormX{\sqity{\qcolorc\qtuq{\color{black}\st\qcolorc}\color{black}}}{\sqity{\tau}}
}

\inferrule[n-rep-conc-tc]{
  \snorm{\st}{\memD}{\memG}{\boxdot; \Upsilon; \Phi}{\sty{\tc}{\sttyidx}}{\memD'}{\memG'}\\\\
  \sty{\tc}{\sttyidx} \leftrightsquigarrow \_ \notin \memD'\\
  \tc \in \boxdot\\
  \tcdef{\tc}{\psi}{\omega} \in \Phi\\
  \keyw{rep}=\stx{rep} \in \omega\\\\
  \snorm{\stx{rep}~\sttyidx}{\memD'}{\cdot}{\boxdot; \emptyset; \Phi}{\sqity{\tau}}{\memD''}{\cdot}
}{
  \snorm{\srep{\st}}{\memD}{\memG}{\boxdot; \Upsilon; \Phi}{\sqity{\tau}}{\memD''}{\memG'}
}

\inferrule[n-rep-conc-parr]{
   \snormX{\st}{\sparr{(\st_1, \st_2)}}\\\\
   \snorm{\srep{\st_1}}{\memD'}{\cdot}{\aCtx}{\sqity{\tau_1}}{\memD''}{\cdot}\\\\
   \snorm{\srep{\st_2}}{\memD''}{\cdot}{\aCtx}{\sqity{\tau_1}}{\memD'''}{\cdot}
}{
   \snorm{\srep{\st}}{\memD}{\memG}{\aCtx}{\sqity{\tau_1 \rightharpoonup \tau_2}}{\memD'''}{\memG'}
}

\inferrule[n-rep-abs]{
  \snorm{\st}{\memD}{\memG}{\boxdot; \Upsilon; \Phi}{\sty{c}{\sttyidx}}{\memD'}{\memG'}\\\\
  \sty{c}{\sttyidx} \leftrightsquigarrow \_ \notin \memD'\\
  c \notin \boxdot\\
  (\alpha~\text{fresh})
}{
  \snorm{\srep{\st}}{\memD}{\memG}{\boxdot; \Upsilon; \Phi}{\sqity{\alpha}}{\memD', \sty{c}{\sttyidx} \leftrightsquigarrow \alpha}{\memG'}
}

\inferrule[n-rep-abs-other]{
  \snormX{\st}{\sotherty{m}{\tau}}\\\\
  \sotherty{m}{\tau} \leftrightsquigarrow \_ \notin \memD'\\
  (\alpha~\text{fresh})
}{
  \snorm{\srep{\st}}{\memD}{\memG}{\aCtx}{\sqity{\alpha}}{\memD', \sotherty{m}{\tau} \leftrightsquigarrow \alpha}{\memG'}
}

\inferrule[n-rep-seen]{
  \snormX{\st}{\st'}\\
  \st' \leftrightsquigarrow \alpha \in \memD'
}{
  \snormX{\srep{\st}}{\sqity{\alpha}}
}

\inferrule[n-qitm-x]{ }{
  \snorm{\sqitm{\qcolor{x}}}{\memD}{\memG}{\aCtx}{\sqitm{x}}{\memD}{\memG}
}

\inferrule[n-qitm-lam]{
  \snorm{\sqity{\qcolor{\tau}}}{\memD}{\memG}{\aCtx}{\sqity{\tau'}}{\memD'}{\memG'}\\\\
  \snorm{\sqitm{\qcolor{\iota}}}{\memD'}{\memG'}{\aCtx}{\sqitm{\iota'}}{\memD''}{\memG''}
}{
  \snorm{\sqitm{\qcolor{\lambda[\tau](x.\iota)}}}{\memD}{\memG}{\aCtx}{\sqitm{\lambda(x.\iota')}}{\memD''}{\memG''}
}
% \inferrule[n-qitm-ap]{
%   \snorm{\sqitm{\qcolor{\iota_1}}}{\memD}{\memG}{\aCtx}{\sqitm{\iota_1'}}{\memD'}{\memG'}\\\\
%   \snorm{\sqitm{\qcolor{\iota_2}}}{\memD'}{\memG'}{\aCtx}{\sqitm{\iota_2'}}{\memD''}{\memG''}
% }{
%   \snorm{\sqitm{\qcolor{\iap{\iota_1}{\iota_2}}}}{\memD}{\memG}{\aCtx}{\sqitm{\iap{\iota_1'}{\iota_2'}}}{\memD''}{\memG''}
% }

\text{\color{gray} (other \qcolor{quoted internal terms} analagously recursive)}

\inferrule[n-qitm-uq]{
  \snormX{\st}{\sqitm{\iota}}
}{
  \snormX{\sqitm{\qcolorc\quq{\color{black}\st\qcolorc}\color{black}}}{\sqitm{\iota}}
}

\inferrule[n-syn]{
  n \hookrightarrow e \in \memG\\
  \esyn{\Upsilon}{\Phi}{e}{\st}{\iota}\\
  (x~\text{fresh})\\\\
  \snorm{\srep{\st}}{\memD}{\cdot}{\boxdot;\Upsilon;\Phi}{\sqity{\tau}}{\memD'}{\cdot}
}{
  \snorm{\ssyn{n}}{\memD}{\memG}{\boxdot; \Upsilon; \Phi}{(\st, \sqitm{x})}{\memD'}{\memG\otimes n \hookrightarrow e : \st \leadsto \iota/x : \tau}
}

\inferrule[n-syn-seen]{
  n \hookrightarrow e : \st \leadsto \iota/x : \tau\in \memG
}{
  \snorm{\ssyn{n}}{\memD}{\memG}{\aCtx}{(\st, \sqitm{x})}{\memD}{\memG}
}

\inferrule[n-syn-fail]{
  n \hookrightarrow e \in \memG\\
  [\Upsilon \vdash_\Phi e \nRightarrow]
}{
  \serr{\ssyn{n}}{\memD}{\memG}{\boxdot; \Upsilon; \Phi}
}

\inferrule[n-ana]{
  \snorm{\st}{\memD}{\memG}{\boxdot;\Upsilon;\Phi}{\st'}{\memD'}{\memG'}\\
  n \hookrightarrow e \in \memG'\\\\
  \eana{\Upsilon}{\Phi}{e}{\st'}{\iota}\\
  (x~\text{fresh})\\
  \snorm{\srep{\st'}}{\memD'}{\cdot}{\boxdot;\Upsilon;\Phi}{\sqity{\tau}}{\memD''}{\cdot}
}{
  \snorm{\sana{n}{\st}}{\memD}{\memG}{\boxdot; \Upsilon; \Phi}{\sqitm{x}}{\memD''}{\memG'\otimes n \hookrightarrow e : \st' \leadsto \iota/x : \tau}
}

\inferrule[n-ana-seen-ok]{
  \snormX{\st}{\st'}\\\\
  n \hookrightarrow e : \st' \leadsto \iota/x : \tau \in \memG'
}{
  \snorm{\sana{n}{\st}}{\memD}{\memG}{\aCtx}{\sqitm{x}}{\memD'}{\memG'}
}

\inferrule[n-ana-seen-err]{
  \snormX{\st}{\st'}\\\\
  n \hookrightarrow e : \st'' \leadsto \iota/x : \tau \in \memG'\\
  \st' \neq \st''
}{
  \serr{\sana{n}{\st}}{\memD}{\memG}{\aCtx}
}

\inferrule[n-ana-fail]{
  \snorm{\st}{\memD}{\memG}{\boxdot;\Upsilon;\Phi}{\st'}{\memD'}{\memG'}\\
  n \hookrightarrow e \in \memG'\\\\
  [\Upsilon \vdash_\Phi e \nLeftarrow \st']
}{
  \serr{\sana{n}{\st}}{\memD}{\memG}{\boxdot; \Upsilon; \Phi}
}

\inferrule[n-raise]{ }{\serrX{\sraise{\kappa}}}
\end{mathpar}
\caption{Static Normalization}
\label{dynamics-SL}
\end{figure*}
%\acks
%The author is grateful to Jonathan Aldrich, Robert Bocchino and anonymous referees for their useful suggestions. This work was funded by the DOE Computational Science Graduate Fellowship under grant number DE-FG02-97ER25308.
%Acknowledgments, if needed.
% \appendix
% \section
% \newpage
% \begin{figure}
% \small
% {\color{gray} (omitted rules generically recursive)} \hfill\fbox{$\kok{\kDelta\minput}{\kTheta\minput}{\kappa\minput}$}
% \begin{mathpar}
% \inferrule[kf-a]{\kalpha \in \kDelta}{\kok{\kDelta}{\kTheta}{\kalpha}}
% ~~~~
% \inferrule[kf-univ]{
%   \kok{\kDelta, \kalpha}{\kTheta}{\kappa}
% }{
%   \kok{\kDelta}{\kTheta}{\kforall{\kalpha}{\kappa}}
% }
% ~~~~
% \inferrule[kf-k]{\kvar \in \kTheta}{\kok{\kDelta}{\kTheta}{\kvar}}
% ~~~~
% \inferrule[kf-ind]{
%   \kok{\kDelta}{\kTheta, \kvar}{\kappa}\\
%   \kpos{\kvar}{\kappa}
% }{
%   \kok{\kDelta}{\kTheta}{\kmu{\kvar}{\kappa}}
% }
% \end{mathpar}
% {\color{gray} (if $\kappa \neq \karrow{\kappa_1}{\kappa_2}$, positivity check is generically recursive)}\hfill\fbox{$\kpos{\kvar\minput}{\kappa\minput}$}
% \begin{mathpar}
% \inferrule[k-arrow-pos]{\kappa_1 \neq \kvar\\\kpos{\kvar}{\kappa_1}\\\kpos{\kvar}{\kappa_2}}{\kpos{\kvar}{\karrow{\kappa_1}{\kappa_2}}}

% \inferrule[k-kvar-pos]{ }{\kpos{\kvar}{\kvar}}
% \end{mathpar}
% ~\hfill\fbox{$\kctxok{\kDelta}{\kGamma}$}\\
% \vspace{-20px}\begin{mathpar}
% \inferrule[kctx-emp]{ }{\kctxok{\kDelta}{\emptyset}}
% ~~~~~~~~~~
% \inferrule[kctx-ext]{
%   \kctxok{\kDelta}{\kGamma}\\
%   \kok{\kDelta}{\emptyset}{\kappa}
% }{\kctxok{\kDelta}{\kGamma, \sx : \kappa}}
% \end{mathpar}
% \caption{Kinds, kinding contexts and equality kinds.}
% \label{statics-SL2}
% \end{figure}
\end{document}

### 2014-07-09 att-popl15/atlam-gensyntax.py

@@ -1,0 +1,109 @@
+def process(filename):
+  with open(filename) as f:
+    grammar_spec_lines = f.readlines()
+
+  syntax = []
+  macros = []
+  n_cols = None
+  n_read = 0
+  read = []
+  for line in grammar_spec_lines:
+    if n_cols is None:
+      n_cols = int(line)
+      continue
+    if line[0:2] == "  ":
+      macros.append(line[2:-1])
+      continue
+    if line == "\n" or line[0] == "%":
+      continue
+    read.append(line[:-1])
+    n_read += 1
+    if n_read == n_cols:
+      syntax.append(read)
+      n_read = 0
+      read = []
+
+  return syntax, macros
+
+ossyntax, osmacros = process("grammar-spec.tex")
+ssyntax, smacros = process("syntax-spec-s.tex")
+qsyntax, qmacros = process("syntax-spec-q.tex")
+esyntax, emacros = process("syntax-spec-e.tex")
+auxsyntax, auxmacros = process("syntax-spec-aux.tex")
+isyntax, imacros = process("syntax-spec-i.tex")
+
+excsyntax, excmacros = process("example-conc.tex")
+exasyntax, examacros = process("example-abs.tex")
+
+def pad_to_same_length(a, b, p):
+  len_difference = len(b) - len(a)
+  if len_difference > 0: # b bigger
+    a.extend([p]*len_difference)
+  elif len_difference < 0: # a bigger
+    b.extend([p]*abs(len_difference))
+
+def combine(left, right):
+  pad_to_same_length(left, right, None)
+  left_n_cols = len(left[0])
+  right_n_cols = len(right[0])
+  combined = [ ]
+  for l, r in zip(left, right):
+    if l is None:
+      l = ["~"]*left_n_cols
+    if r is None:
+      r = ["~"]*right_n_cols
+    c = list(l)
+    c.extend(r)
+    combined.append(c)
+  return combined
+
+syntax_SL = ssyntax
+syntax_OSL = combine(ossyntax, qsyntax)
+syntax_EL = esyntax#combine(esyntax, auxsyntax)
+syntax_IL = isyntax
+example = combine(excsyntax, exasyntax)
+smacros.extend(qmacros)
+smacros.extend(emacros)
+smacros.extend(auxmacros)
+smacros.extend(imacros)
+smacros.extend(osmacros)
+smacros.extend(excmacros)
+smacros.extend(examacros)
+macros = smacros
+
[…]

### 2014-07-22 atlang-gpce14/atlang-gpce14.tex
GPCE14 post-mortem

@@
 
-\lefthyphenmin=62
+\lefthyphenmin=8
 \sloppy
@@
 %abstractions, all within libraries.
-Programmers justifiably prefer abstractions available as libraries for an existing language over those requiring a switch to a new language, but library providers cannot extend these language's type systems directly, so powerful statically typed abstractions often remain obscure.
-This suggests a need for a language that is \emph{backwards compatible} with the libraries, tools and infrastructure of a widely adopted language but that has an  \emph{internally extensible type system}, organized around a mechanism that guarantees that any combination of extensions can be imported  without  ambiguity.
-
-Our aim in this paper is thus to 1) introduce such a mechanism and 2) show how it can be implemented in a backwards compatible manner within an existing widely used language.
-We do so with two artifacts: @\texttt{lang}, an  extensible statically typed language implemented as a library inside Python, and @$\lambda$, a minimal calculus.
+Programmers justifiably prefer abstractions available as libraries for the language they already use over those requiring a switch to a new language.
+But library providers cannot typically extend the type systems of these languages directly, so powerful statically typed abstractions  remain obscure.
+This suggests that a language is needed that is \emph{backwards compatible} with the libraries and tools of a widely adopted language, but that has an  \emph{internally extensible type system}, organized around a mechanism that guarantees that any combination of extensions can be imported  without  ambiguity.
+
+Our aim in this paper is to 1) introduce such a mechanism, and 2) show how it can itself be implemented as a library for an existing widely used language.
+We describe two artifacts: @\texttt{lang}, an  extensible statically typed language implemented as a library inside Python, and @$\lambda$, a minimal core calculus.
 Both @\texttt{lang} and @$\lambda$ have a fixed syntax but  determine the semantics of each term constructor by delegating control, according to a novel bidirectional typechecking and translation protocol, to methods associated with a relevant \emph{active type constructor}, defined in a library.
-The type constructor delegated control over a term is stable even when additional extensions are imported, so we avoid ambiguities by construction.
-% By flexibly  repurposing a fixed syntax in this type-directed manner, we avoid ambiguities by construction.
+The delegate is stable even when additional extensions are imported, so we avoid ambiguities by construction.
+We demonstrate with several examples of interesting type system fragments as libraries.% By flexibly  repurposing a fixed syntax in this type-directed manner, we avoid ambiguities by construction.
 %In @\texttt{lang}, types and functions are defined using Python's syntax, distributed using Python's package system and can call into Python libraries, so @\texttt{lang} is uniquely backwards compatible.
@@
 \section{Introduction}\label{intro}
-Asking programmers to import a library is simpler than asking them to adopt a new programming language.
+Asking programmers to import a library is easier than asking them to adopt a new programming language.
 Indeed, recent empirical studies underscore the difficulties of driving new languages into adoption, finding that extrinsic factors like compatibility with large existing code bases, library availability, familiarity and tool support are at least as important as intrinsic features of the language \cite{Meyerovich:2013:EAP:2509136.2509515,chen05}.
 
-Unfortunately, researchers and domain experts designing new abstractions can  find it difficult to soundly and naturally implement them as libraries, particularly when these require strengthening the language's static type system.
+Unfortunately, researchers and domain experts designing new abstractions can  find it difficult to soundly and naturally implement them as libraries, particularly when they require strengthening the language's static type system.
 In these situations, abstraction providers generally develop a new language dialect, increasingly with the aid of tools like language workbenches \cite{erdweg2013state}.
@@
 \inferrule[att-ana-lam]{
-    \sigma_1~\mathtt{ty}_\fvalCtx\\\sigma_2~\mathtt{ty}_\fvalCtx\\
+   \Phi_\text{fn} \subset \Phi\\ \sigma_1~\mathtt{ty}_\fvalCtx\\\sigma_2~\mathtt{ty}_\fvalCtx\\
   \Gamma, x \Rightarrow \sigma_1 \vdash_\fvalCtx e \Leftarrow \sigma_2 \leadsto \iota
 }{
-  \atjanaX{\lambda x.e}{\FF{ty}[\fvar{fn}](\FF{cons}(\sigma_1; \FF{cons}(\sigma_2; \FF{nil})))}{\lambda(x.\iota)}
+  \atjanaX{\lambda(x.e)}{\FF{ty}[\fvar{fn}](\FF{cons}(\sigma_1; \FF{cons}(\sigma_2; \FF{nil})))}{\lambda(x.\iota)}
 }
@@
 \inferrule[att-syn-idx-lam]{
-    \sigma_1~\mathtt{ty}_\fvalCtx\\
+   \Phi_\text{fn} \subset \Phi\\ \sigma_1~\mathtt{ty}_\fvalCtx\\
   \Gamma, x \Rightarrow \sigma_1 \vdash_\fvalCtx e \Rightarrow \sigma_2 \leadsto \iota

### 2014-08-01 sanitation-psp14/sanitation-psp14.tex
Added PSP changes.

@@
-\documentclass[10pt,preprint]{sigplanconf}
+\documentclass{acm_proc_article-sp}
 \newcommand{\lamAce}{\lambda_{\text{Ace}}}
@@
 \usepackage{amssymb}
+
+% Hack. Something's wrong with PLAS paper when using the ACM Proc docclass
+\let\proof\relax
+\let\endproof\relax
 \usepackage{amsthm}
+
 \usepackage{ stmaryrd }
@@
 % Source language terms.
-\newcommand{\sisubst}[3]{{\sf rreplace}[#1](#2,#3)}
+\newcommand{\sisubst}[3]{{\sf rreplace}[#1](#2,#3)} \newcommand{\rreplace}[3]{{\sf rreplace}[#1](#2,#3)} % lots of legacy naming around these parts...
+
 \newcommand{\rssreplace}[3]{\sisubst{#1}{#2}{#2}} % TODO-nrf fix this.
@@
 \newcommand{\coerce}[2]{ {\sf rcoerce}[#1](#2)}
-\newcommand{\sistr}[1]{{\sf rstr}[#1]}
+\newcommand{\sistr}[1]{{\sf rstr}[#1]}   \newcommand{\rstr}[1]{{\sf rstr}[#1]} % Lots of legacy naming around these parts...
+
 \newcommand{\strin}[1]{\sistr{#1}}
@@
 % target language terms.
-\newcommand{\tsubst}[3]{{\sf replace}(#1,#2,#3)} % TODO-nrf rename the commands.
+\newcommand{\tsubst}[3]{{\sf replace}(#1,#2,#3)} \newcommand{\metareplace}[3]{{\sf replace}(#1,#2,#3)} % TODO-nrf rename the commands. Lots of legacy naming around these parts...
+
 \newcommand{\tcheck}[2]{{\sf check}(#1, #2)}
@@
 
-\conferenceinfo{-}{-} 
-\copyrightyear{-} 
-\copyrightdata{[to be supplied]}
+%dep_docclass\conferenceinfo{-}{-} 
+%dep_docclass\copyrightyear{-} 
+%dep_docclass\copyrightdata{[to be supplied]}
 
@@
 %\authorinfo{~}{~}{~}
-\authorinfo{Nathan Fulton\and Cyrus Omar\and Jonathan Aldrich}
-           {School of Computer Science\\
-            Carnegie Mellon University}
-           {\{nathanfu, comar, aldrich\}@cs.cmu.edu}
+%dep_docclass\authorinfo{Nathan Fulton\and Cyrus Omar\and Jonathan Aldrich}
+%dep_docclass           {School of Computer Science\\
+%dep_docclass            Carnegie Mellon University}
+%dep_docclass           {\{nathanfu, comar, aldrich\}@cs.cmu.edu}
 
@@
 
+there's extra space here because of the next page...
+
 %The $\lcs$ language gives static semantics for common regular expression library
@@
 
+% I can't achieve the correct latout here. We should have something like:
+%       Lam_S Synax         Lam_P Syntax
+%       Lam_S Sematics      LAm_P Semantics
+%                Translation rules
+% Or, alternatively, we can flip reflect things so that the entire page
+% reads top-down (Lam_S syntax Lam_S smantics horizontally aligned w/ one
+% another).
+
+% Unfortunately, I can't get that to work. So, we're settling for something
+% slightly ugly, where the fitures appear in the incorrect order and the 
+% translation rules are on top.
+
+% The translations rules *just won't* go to the bottom of the page, so 
+% instead I am placing them at the top. Since the figures are out of order,
+% it's even more important that we get the number gith. therefore, the counter
+% is modified before and after this figure.
+\addtocounter{figure}{4}
+\begin{figure*}
+
+$\fbox{\inferrule{}{\trden{S} = P}}$
+\begin{mathpar}
+
+\inferrule[Tr-string]
+{ \ }
+{\trden{\strin{s}} = \tstr{s}}
+
+\inferrule[Tr-Concat]
+{ \trden{S_1} = P_1 \\ \trden{S_2} = P_2}
+{\trden{\rsconcat{S_1}{S_2}} = \tconcat{P_1}{P_2}}
+
+\inferrule[Tr-Subst]
+{ \trden{S_1} = P_1 \\ \trden{S_2} = P_2 }
+{\trden{ \sisubst{r}{S_1}{S_2} } = \tsubst{\rx{r}}{P_1}{P_2} }
+
+\inferrule[Tr-Coerce-Ok]
+{ S : \strin{r} \\ \lang{r'} \subseteq \lang{r}}
+{\trden{ \coerce{r'}{S} } = \tstr{s}}
+
+\inferrule[Tr-Coerce-NotOk]
+{\trden{S} = P \\ S : \strin{r} \\ \lang{r'} \not \subseteq \lang{r}}
+{\trden{\coerce{r'}{S}} = \tcheck{\rx{r'}}{P}}
+
+\end{mathpar}
+\caption{Translation from source terms (S) to target terms (P).
+The translation is type-directed in the Tr-Coerce cases.}
+\end{figure*}
+\addtocounter{figure}{-5}
+
 %
@@
 \newcommand{\ereduces}{\Downarrow}
-
-\afterpage{
 
@@
 {\tctx \vdash \tcheck{P_1}{P_2} : \str}
-
 \end{mathpar}
@@
 
-\clearpage
-} % End \afterpage
-
 \newpage
-
-%
-% Begin translation
-%
-
-\begin{figure}[t]
-
-$\fbox{\inferrule{}{\trden{S} = P}}$
-\begin{mathpar}
-
-\inferrule[Tr-string]
-{ \ }
-{\trden{\strin{s}} = \tstr{s}}
-
-\inferrule[Tr-Concat]
-{ \trden{S_1} = P_1 \\ \trden{S_2} = P_2}
-{\trden{\rsconcat{S_1}{S_2}} = \tconcat{P_1}{P_2}}
-
-\inferrule[Tr-Subst]
-{ \trden{S_1} = P_1 \\ \trden{S_2} = P_2 }
-{\trden{ \sisubst{r}{S_1}{S_2} } = \tsubst{\rx{r}}{P_1}{P_2} }
-
-\inferrule[Tr-Coerce-Ok]
-{ S : \strin{r} \\ \lang{r'} \subseteq \lang{r}}
-{\trden{ \coerce{r'}{S} } = \tstr{s}}
-
-\inferrule[Tr-Coerce-NotOk]
-{\trden{S} = P \\ S : \strin{r} \\ \lang{r'} \not \subseteq \lang{r}}
-{\trden{\coerce{r'}{S}} = \tcheck{\rx{r'}}{P}}
-
-\end{mathpar}
-\caption{Translation from source terms (S) to target terms (P).
-The translation is type-directed in the Tr-Coerce cases.}
-\end{figure}
-
+\ 
+\newpage
+\ 
+\newpage
 Note that we never insert an internal check where the type of the string implies
@@
 \begin{thm}
-  Let $S$ be a term in the source language. If $\sctx \vdash S : \psi$ 
-  then $S \sreduces S'$ and $S' : \psi$.
+  Let $S$ be a term in the source language. If $\sctx \vdash S : \stringin{r}$ 
+  then $S \sreduces \rstr{s}$ and $\rstr{s} : \stringin{r}$, or else $S \sreduces \err$.
 \end{thm}
 \begin{proof}
-  TODO-nrf
-
-By a straightforward induction on the derivation of $\sctx \vdash S : \psi$.
-Note that preservation depends upon the substitution correspondence theorem
-for the T-T-Subst case and the closure lemmas for the T-T-Concat case.
+By induction on the derivation of $\sctx \vdash S : \psi$. The interesting case
+is S-T-Replace, which requires Lemma C.
+
+\begin{description}
+\item[S-T-Stringin-I:]
+If $S = \rsconcat{S_1}{S_2} : \stringin{r}$ then
+$S \treduces S$ by S-E-RStr, and $\sctx \vdash S : \psi$ by assumption.
+
+\item[S-T-Concat:] 
+Suppose $S = \rsconcat{S_1}{S_2} : \stringin{r_1 r_2}$.
+By inversion, $\sctx \vdash S_1 : \stringin{r_1}$ and $\sctx \vdash S_1 : \stringin{r_2}$.
+It follows by induction that either $S_1 \treduces \err$, $S_2 \treduces \err$, 
+or $S_1 \treduces \rstr{s_1}$ and $S_2 \treduces \rstr{s_2}$ for some $s_1 \in \lang{r_1}$ and $s_2 \in \lang{r_2}$.
+In the latter case $S \reduces \rstr{s_1s_2}$ by S-E-Concat and $\sctx \vdash \rstr{s_1s_2} : \tstr{r_1r_2}$ by Lemma B.
+In the former cases, $S \sreduces \err$.
+
+\item[S-T-Replace]:
+Suppose $S = \rreplace{r}{S_1}{S_2}$ and $\sctx \vdash S : \stringin{r'}$.
+By inversion $\sctx \vdash S_1 : \stringin{r_1}$ and $\sctx \vdash S_2 : \stringin{r_2}$ such that $\lsubst{r}{r_1}{r_2} = r'$.
+By induction, $S_1 \sreduces \err$, $S_2 \sreduces \err$ 
+or $S_1 \sreduces \rstr{s_1}$ and $S_2 \sreduces \rstr{s_2}$ such that 
+In the latter case, we know $\metareplace{r}{s_1}{s_2} \in \lang{\lsubst{r}{r_1}{r_2}}$ by Lemma C;
+therefore by S-E-Replace, $S \treduces \rstr{s}$ such that  $s \in \lang{\lsubst{r}{r_1}{r_2}} = \lang{r'}$.
+So by S-T-String-I, $\rstr{s} : \stringin{r'}$.
+In the former cases, $S \treduces \err$.
+
+\item[S-T-Coerce]:
+Suppose $S = \coerce{r}{S_1}$ and $S : \stringin{r}$.
+By inversion, $\sctx \vdash S_1 : \stringin{r'}$.
+By induction, $S_1 \treduces \err$ or $S_1 \treduces \rstr{s}$.
+In the former case $S \treduces \err$ by propagation rules.
+In the latter case we have by Lemma B(2) that $s \in \lang{r}$ or else
+$s \not \in \lang{r}$.
+If $s \in \lang{r}$ then $\rstr{s} : \stringin{r}$.
+If $s \not \in \lang{r}$ then $S \treduces \err$. 
+\end{description}
 \end{proof}
@@
 \begin{proof}
-TODO-nrf
-
-  By a straightforward induction on the derivation of $\tctx \vdash P : \theta$.
-Note that the well-formedness of regular expressions indexing ${\tt rx}$ types
-depends upon the closure lemmas.
+The proof proceeds by induction on the typing relation and is trivial give
+and inversion lemma for the typing relation. \texbf{We can write up this proof if we end up having enough space...}
 \end{proof}

### 2014-08-01 sanitation-psp14/sanitation-psp14.tex
deprecated old confusing proof.

@@
 The proof proceeds by induction on the typing relation and is trivial give
-and inversion lemma for the typing relation. \texbf{We can write up this proof if we end up having enough space...}
+and inversion lemma for the typing relation. \textbf{We can write up this proof if we end up having enough space...}
 \end{proof}
@@
 \begin{proof}
-By induction on the derivation of $\trden{S} = T$. TODO-nrf add case magic.
+  \textbf{This is a proof for a different theorem that was swapped out at some point, 
+    and is no longer valid. Working on the new proof now, by induction on the typing
+  relation and correct choice of T}.
+  By induction on the derivation of $\trden{S} = T$. (is this an old proof?) 
 \begin{description}

### 2014-08-04 sanitation-psp14/sanitation-psp14.tex
S/P/trans version

--- 2014-08-01 sanitation-psp14/sanitation-psp14.tex
@@
 \newcommand{\coerce}[2]{ {\sf rcoerce}[#1](#2)}
+\newcommand{\rcoerce}[2]{{\sf rcoerce}[#1](#2)}
 \newcommand{\sistr}[1]{{\sf rstr}[#1]}   \newcommand{\rstr}[1]{{\sf rstr}[#1]} % Lots of legacy naming around these parts...
@@
 \newcommand{\strin}[1]{\sistr{#1}}
-\newcommand{\rsconcat}[2]{{\sf rconcat}(#1,#2)}
+\newcommand{\rsconcat}[2]{{\sf rconcat}(#1,#2)} \newcommand{\rconcat}[2]{{\sf rconcat}(#1,#2)} % lots of legact naming around these parts..
 
@@
 \renewcommand{\tstr}[1]{{{\sf str}[#1]}}
-\newcommand{\tconcat}[2]{{\sf concat}(#1,#2)}
+\newcommand{\preplace}[3]{{\sf preplace}(#1,#2,#3)}
+\newcommand{\tconcat}[2]{{\sf concat}(#1,#2)} \newcommand{\concat}[2]{{\sf concat}(#1,#2)} % lots of legacy naming around these parts...
+
 \newcommand{\regext}[1]{ {\sf rx}[#1] } % TODO-nrf remove
@@
 \newcommand{\dosubst}[3]{{\tt replace}(#1,#2,#3)}
-\newcommand{\lsubst}[3]{{\tt lreplace}(#1,#2,#3)}
+\newcommand{\lsubst}[3]{{\tt lsubst}(#1,#2,#3)} % This used to render lreplace(...) so there're probably mistkes wherever this command was used now.
+\newcommand{\lreplace}[3]{{\tt lreplace}(#1, #2, #3)}
 
@@
 casting.
-%
+
+Note that we never insert an internal check where the type of the string implies
+that a check must succeed. Furthermore, expensive calculations (such as language
+inclusion or ${\tt dsubst}$, all occur at compile time.
+Since sanitation problems are generally expressible with smalle, simple expressions,
+we believe that the compile-time overhead is not significant enough to prohibiour target usecase. Informal experimentation with our implementation seems to
+support this assertion.
+
+Due to Ei-Replace, all of our meta-theory depends upon a correct implementation
+of both regular expression replacement and inclusion checking in the internal language.
+We believe this assumption is okay for two reasons. First, our system is still
+superior to the status quo, which relies upon the correctness of these libraries
+\emph{in addition to} correct application logic. Second, regular expression libraries are
+generally well-tested and there exist verified implementations.
+
 % Begin Grammar 
-%
-
-% Remove the < > from grammar productions.
+% Removes the < > from grammar productions.
 \renewcommand{\grammarlabel}[2]{#1\hfill#2}
 
-\begin{figure}
+\begin{figure}[h]
 \begin{grammar}
@@
 
-\begin{figure}
+\begin{figure}[h]
 \begin{grammar}
@@
 {\tctx \vdash P_1 : \regex \\ \tctx \vdash P_2 : \str \\ \tctx \vdash P_3 : \str }
-{\tctx \vdash \tsubst{P_1}{P_2}{P_3} : \str}
+{\tctx \vdash \preplace{P_1}{P_2}{P_3} : \str}
 
@@
 {P_1 \treduces \rx{r} \\ P_2 \treduces \tstr{s_2} \\ P_3 \treduces \tstr{s_3} \\ \dosubst{r}{s_2}{s_3} = s} 
-{\tsubst{P_1}{P_2}{P_3} \treduces \tstr{s}}
+{\preplace{P_1}{P_2}{P_3} \treduces \tstr{s}}
 
@@
 \newpage
-Note that we never insert an internal check where the type of the string implies
-that a check must succeed. Furthermore, expensive calculations (such as language
-inclusion or ${\tt dsubst}$, all occur at compile time.
-Since sanitation problems are generally expressible with smalle, simple expressions,
-we believe that the compile-time overhead is not significant enough to prohibiour target usecase. Informal experimentation with our implementation seems to
-support this assertion.
-
-Due to Ei-Replace, all of our meta-theory depends upon a correct implementation
-of both regular expression replacement and inclusion checking in the internal language.
-We believe this assumption is okay for two reasons. First, our system is still
-superior to the status quo, which relies upon the correctness of these libraries
-\emph{in addition to} correct application logic. Second, regular expression libraries are
-generally well-tested and there exist verified implementations.
 
@@
 
-The soundness theorems in this section depend upon some well-known properties
-of regular expressions. Throughout this section, suppose regular expression
-$r$ and strings $s$ are defined over a single, fixed, finite alphabet $\Sigma$.
-
-\begin{thm} \label{thm:closure}
-The regular expressions are closed under complements and concatenation.
-\end{thm}
-\begin{proof}
-See \cite{cinderella}.
-\end{proof}
-
-\begin{defn}[$\tt{subst}$]
-  The function $\subst{r}{s_1}{s_2}$ produces a string in which all substrings
-  of $s_1$ matching $r$ are replaced with $s_2$.
+Our type safety proofs for languages S and P and our translation correctness
+result all depend on some properties of regular languages. The crucial property
+is a relationship between string substitution -- which is available in any regular
+expression library -- and regular language substitution, which is a corresponding operation
+on languages instead of strings \ref{thm:substcorrespondence}.
+The decidability of language substitution is what enables static analysis of 
+sanitation algorithms implemented in terms of string replacement
+
+Throughout this section, we fix an alphabet $\Sigma$ over which strings $s$ and
+regular expressions $r$ are defined. throughout the paper, $\lang{r}$ refers to the
+language recognized by the expression $r$. This distinction between the expression
+and its language -- typically elided in the literature -- makes our definition and
+proofs about systems S and P more readable.
+
+\begin{lem}{Properties of Regular Languages and Expressions.} \label{thm:regexprops}
+The following are well-known properties of regular expressions which are necessary for our proofs:
+\begin{description}
+\item[(1):] If $s_1 \in \lang{r_1}$ and $s_2 \in \lang{r_2}$ then $s_1s_2 \in \lang{r_1r_2}$
+\item[(2):] For all strings $s$ and expressions $r$, either $s \in \lang{r}$ or $s \not \in \lang{r}$.
+\end{description}
+\item[(3):] Regular languages are closed under complements and concatenation.
+\item[(4):] The regular expressions correspond bijectively to the regular languages.
+\end{lem}
+
+\begin{defn}[$\tt{lsubst}$]
+The function $\lsubst{r}{s_1}{s_2}$ produces a string in which all substrings of $s_1$ matching $r$ are replaced with $s_2$.
 \end{defn}
 
-\begin{defn}[$\tt{lsubst}$]
-  The function $\lsubst{r}{r_1}{r_2}$ produces a regular expression in which
-  and sublanguage $r_1'$ of $r_1$ satisfying the condition $\lang{r_1'} \subseteq{r}$
-  is replaced with $r_2$.
+\begin{defn}[$\tt{lreplace}$]
+The function $\lreplace{r}{r_1}{r_2}$ produces a regular expression in which any sublanguage $\lang{r_1'}$ of $\lang{r_1}$ satisfying the condition $\lang{r_1'} \subseteq{\lang{r}}$ is replaced with $\lang{r_2}$.
 \end{defn}
 
-We now present some properties about substitution.
-
-\begin{thm}
-  String substitution ($\subst{r}{s_1}{s_2}$) is decidable and defined for all expressions
-  and strings over similar alpha.
-\end{thm}
-
-\begin{thm}
-  Language substittion $(\lsubst{r}{r_1}{r_2})$ is decidable and defined for
-  all expressions.
-\end{thm}
-
-The following relationship between string substituion and language substitution is at
-the heart of our main correctness reesult.
-
-\begin{thm}
-  For all $r$, if $s_1 \in r_1$ and $s_2 \in r_2$ then 
-  $\subst{r}{s_1}{s_2} \subseteq \lsubst{r}{r_1}{r_2}$.
-\end{thm}
-
-\begin{lem}{Coercion Lemma.} \label{thm:coercionalemma}
-Suppose that $R$ and $L$ are regular expressions, and that $s \in R$ is a finite string.
-Let $s' := coerce(R,L,s)$ with all maximal substrings recognized by $L$ replaced with $\epsilon$.
-Then $s'$ is recognized by $(R \backslash L) + \emptyset$ and the construction of $R \backslash L$ is decidable.
+\begin{lem}{Closure and Totality of Replacement.} \label{thm:total}
+If $r, r_1$ and $r_2$ are regular expressions, then $\lreplace{r}{r_1}{r_2}$ is also a regular expression.
 \end{lem}
 \begin{proof}
-Let $F,G$ be FAs corresponding to $R$ and $L$, and let $G'$ be $G$ with its final states inverted (so that $G'$ is the complement of $L$).
+TODO-nrf 
+\end{proof}
+
+\begin{lem}{Substitution Correspondence.} \label{thm:substcorrespondence}
+If $s_1 \in \lang{r_1}$ and $s_2 \in \lang{r_2}$ then $\lsubst{r}{s_1}{s_2} \in \lreplace{r}{r_1}{r_2}$.
+\end{lem}
+\begin{proof}
+TODO-nrf 
+\end{proof}
+
+% NRF The coercion lemma is no longer necessary, because it is a property of
+% "strip"-style sanitation rather than the more general case of replacement 
+% treated in this paper. However, we should mention this difference between my
+% undergraduate thesis and this paper in the related work section if we are
+% going tin clude a discussio of unpublished or quasi-published (e.g. src)
+% results.
+%\begin{lem}{Coercion Lemma.} \label{thm:coercionalemma}
+%Suppose that $R$ and $L$ are regular expressions, and that $s \in R$ is a finite string.  
+%Let $s' = \coerce(R,L,s)$ with all maximal substrings recognized by $L$ replaced with $\epsilon$.
+Then $s'$ is recognized by $(R \backslash L) + \emptyset$ and the construction of $R \backslash L$ is decidable.
+%\end{lem}
+%\begin{proof}
+%Let $F,G$ be FAs corresponding to $R$ and $L$, and let $G'$ be $G$ with its final states inverted (so that $G'$ is the complement of $L$).
 Define an FA $H$ as a DFA corresponding to the NFA found by combining $F$ and $G'$ such that $H$ accepts only if $R$ and $L'$ accept or if $s$ is empty (this construction may result in an exponential blowup in state size.)  Clearly, $H$ corresponds to $R \backslash L + \emptyset$.
 Thus, the construction of $R \backslash L + \emptyset$ is decidable.
-
-If $R \subset L$, $s' = \emptyset$.
+%
+%If $R \subset L$, $s' = \emptyset$.
 If $L \subset R$, either $s' = \emptyset$, or $s' \in R$ and $s' \not \in L$.
@@
 The subset language is considered above and the disjoint language is inconsequential.
-\end{proof}
+%\end{proof}
+
+% Why is this necessary?
+%\begin{lem}
+%  For all $r$, if $s_1 \in r_1$ and $s_2 \in r_2$ then $\subst{r}{s_1}{s_2} \subseteq \lsubst{r}{r_1}{r_2}$.
+%\end{lem}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+% BEGIN PROOFS ABOUT S AND P
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 
@@
 \begin{lem}
-  If $\sctx \vdash S : \stringin[r]$ then $r$ is a well-formed regular expression. 
+  If $\sctx \vdash S : \stringin{r}$ then $r$ is a well-formed regular expression. 
 \end{lem}
 \begin{proof}
-  By induction on the typing derivation and the closure lemmas for regular
-  expressions.
+  The only non-trivial case is S-T-Replace, which follows from \ref{thm:total}. 
 \end{proof}
 
+\begin{lem}
+  If $\tctx \vdash P : \regex$ then $P \treduces \rx{r}$ such that $r$ is a well-formed regular expression. 
+\end{lem}
+
 We now prove safety for the string fragment of the source and target languages.
 
-\begin{thm}
+\begin{thm}{Safety for the String Fragment of P.}
   Let $S$ be a term in the source language. If $\sctx \vdash S : \stringin{r}$ 
@@
 If $S = \rsconcat{S_1}{S_2} : \stringin{r}$ then
-$S \treduces S$ by S-E-RStr, and $\sctx \vdash S : \psi$ by assumption.
+$S \sreduces S$ by S-E-RStr, and $\sctx \vdash S : \psi$ by assumption.
 
@@
 By inversion, $\sctx \vdash S_1 : \stringin{r_1}$ and $\sctx \vdash S_1 : \stringin{r_2}$.
-It follows by induction that either $S_1 \treduces \err$, $S_2 \treduces \err$, 
-or $S_1 \treduces \rstr{s_1}$ and $S_2 \treduces \rstr{s_2}$ for some $s_1 \in \lang{r_1}$ and $s_2 \in \lang{r_2}$.
-In the latter case $S \reduces \rstr{s_1s_2}$ by S-E-Concat and $\sctx \vdash \rstr{s_1s_2} : \tstr{r_1r_2}$ by Lemma B.
+It follows by induction that either $S_1 \sreduces \err$, $S_2 \sreduces \err$, 
+or $S_1 \sreduces \rstr{s_1}$ and $S_2 \sreduces \rstr{s_2}$ for some $s_1 \in \lang{r_1}$ and $s_2 \in \lang{r_2}$.
+In the latter case $S \reduces \rstr{s_1s_2}$ by S-E-Concat and $\sctx \vdash \rstr{s_1s_2} : \tstr{r_1r_2}$ by \ref{thm:regexprops}.
 In the former cases, $S \sreduces \err$.
 
-\item[S-T-Replace]:
+\item[S-T-Replace:]
 Suppose $S = \rreplace{r}{S_1}{S_2}$ and $\sctx \vdash S : \stringin{r'}$.
@@
 or $S_1 \sreduces \rstr{s_1}$ and $S_2 \sreduces \rstr{s_2}$ such that 
-In the latter case, we know $\metareplace{r}{s_1}{s_2} \in \lang{\lsubst{r}{r_1}{r_2}}$ by Lemma C;
-therefore by S-E-Replace, $S \treduces \rstr{s}$ such that  $s \in \lang{\lsubst{r}{r_1}{r_2}} = \lang{r'}$.
+In the latter case, we know $\lreplace{r}{s_1}{s_2} \in \lang{\lsubst{r}{r_1}{r_2}}$ by Lemma C;
+therefore by S-E-Replace, $S \sreduces \rstr{s}$ such that  $s \in \lang{\lsubst{r}{r_1}{r_2}} = \lang{r'}$.
 So by S-T-String-I, $\rstr{s} : \stringin{r'}$.
-In the former cases, $S \treduces \err$.
-
-\item[S-T-Coerce]:
+In the former cases, $S \sreduces \err$.
+
+\item[S-T-Coerce:]
 Suppose $S = \coerce{r}{S_1}$ and $S : \stringin{r}$.
 By inversion, $\sctx \vdash S_1 : \stringin{r'}$.
-By induction, $S_1 \treduces \err$ or $S_1 \treduces \rstr{s}$.
-In the former case $S \treduces \err$ by propagation rules.
-In the latter case we have by Lemma B(2) that $s \in \lang{r}$ or else
+By induction, $S_1 \sreduces \err$ or $S_1 \sreduces \rstr{s}$.
+In the former case $S \sreduces \err$ by propagation rules.
+In the latter case we have by property 2 of \ref{thm:regexprops} that $s \in \lang{r}$ or else
 $s \not \in \lang{r}$.
 If $s \in \lang{r}$ then $\rstr{s} : \stringin{r}$.
-If $s \not \in \lang{r}$ then $S \treduces \err$. 
+If $s \not \in \lang{r}$ then $S \sreduces \err$. 
 \end{description}
@@
 \begin{thm}
-If $S : \strin{r}$ then there exists a T such that $\trden{s} = T$ and either: 
+If $S : \strin{r}$ then there exists a P such that $\trden{s} = P$ and either: 
 \begin{description}
-  \item[(a)] $T \treduces \tstr{s}$ and $S \sreduces \strin{s}$, and $s \in lang{r}$.
-  \item[(b)] $T \err$ and $S \err$.
+  \item[(a)] $P \treduces \tstr{s}$ and $S \sreduces \strin{s}$, and $s \in lang{r}$.
+  \item[(b)] $P \err$ and $S \err$.
   \end{description}
 \end{thm}
+
 \begin{proof}
-By induction on the derivation of $\trden{S} = T$. TODO-nrf add case magic.
+The proof proceeds by induction on the typing relation for $S$.
+Throughout the proof, properties from the closure lemma for regular languages
+are necessary; for brevity, we elide these references.
+
 \begin{description}
-\item[Tr-string] By assumption $\trden{S} = \tstr{s}$. The (a) property follows.
-By S-E-Str, $S \sreduces \str{s}$ and by T-E-Str, $\tstr{s} \treduces \tstr{s}$.
-By source language type safety, it follows that $s \in \lang{r}$.
-\item[Tr-Concat]
-By inversion of the typing relaiton for the source lnaguage, $S_1 : \strin{r_1}$ and $S_2 : \stringin{S_2}$.
-By induction on the left premise, $T_1 \treduces \tstr{s_1}$, $S_1 \sreduces \str{s_1}$ and $s_1 \in \lang{r_1}$
-or else $T_1 err$. 
-Similarly for $T_2$ and $S_2$.
-In the error cases, the propagations rules corresponding to S-E-Concat and T-E-Concat imply that $\tconcat{T_1}{T_2} \err$ and $\rsconcat{S_1}{S_2} \err$.
-In the non-error case, $\rsconcat{S_1}{S_2} \sreduces \strin{s_1 s_2}$ by S-E-Concat and $\tconcat{T_1}{T_2} \treduces \tstr{s_1 s_2}$ by T-E-Concat.
-Finally, $s_1 \in \lang{r_1}$ and $s_2 \in \lang{r_2}$ implies that $s_1 s_2 \in \lang{r}$ by \ref{lemmaregex}.
-\item[Tr-subst]
-Suppose $\sisubst{r}{S_1}{S_2} : \strin{s}$ for some $x$. 
-By inverison, $S_1 : \strin{r_1}$ and $S_2 : \strin{r_2}$.
-Note that $s : \subst{r}{S_1}{S_2} : \stringin{\tsubst{r}{r_1}{r_2}}$ by S-T-Subst.
-By induction, $\trden{S_1} = T_1$ and $\trden{S_2} = T_2$.
-Either (a) or (b) holds for each.
-
-If (a) holds for both, the $S_1 \sreduces \strin{S_1}$ and $S_2 \sreduces \strin{S_2}$ for $s_1 \in \lang{r_1}$ and $s_2 \in \lang{r_2}$.
-Therefore, $\sisubst{r}{S_1}{S_2} \sreduces \strin{\dosubst{r,s_1,s_2}}$.
-The same fact can be shown in an analogous manner for $T_1$ and $T_2$.
-Finally, $s_1 \in \lang{r_1}$ and $s_2 \in \lang{r_2}$, which implies that $\dosubst{r}{s_1}{s_2} \in \lang{\sisubst{r}{s_1}{s_2}}$ by \ref{substhm}.
-
-If (b) holds for either the left or right premise, then $\sisubst{r}{S_1}{S_2} \err$ by the propagation rules corresponding to S-T-subst, 
-and $\tsubst{S_1}{S_2} \err$ by the propagation rules corresponding to T-T-Subst.
-
-\item[Tr-Coerce-Ok] TODO-nrf.
-
-\item[Tr-Coerce-NotOk] TODO-nrf.
+\item[S-T-String-I:] 
+Let $S = \rstr{s}$ and suppose $\sctx \vdash \rstr{s} : \stringin{r}$.
+Choose $T = \str{s}$ and note that $\trden{S} = P$ by Tr-String.
+By P-E-String, $P \treduces \str{s}$ and by S-E-String $S \treduces \rstr{s}$.
+Finally, by inversion of S-T-Stringin-I, $s \in \lang{r}$.
+
+\item[S-T-Concat:]
+Let $S = \rconcat{S_1}{S_2}$ and suppose $\sctx \vdash S : \stringin{r_1 r_2}$.
+By inversion, $\sctx \vdash S_1 : \stringin{r_1}$.
+It follows by induction that there exists a $P_1$ such that $\trden{S_1} = P_1$.
+By a similar argument for $S_2$ and $r_2$, there exists a $P_2$ such that $\trden{S_2} = P_2$.
+Choose $P = \concat{P_1}{P_2}$.
+
+We first prove property (a).
+Note that $S_1$ and $P_1$ are well typed (nrf ACTUALLY WE DON'T KNOW THAT $P_1$ IS WELL-TYPED!) and do not result in errors.
+Therefore, $S_1 \sreduces \rstr{s_1}$ and $P_1 \treduces \str{s_1}$ for some $s_1 \in \lang{r_1}$ by theorems 8 and 9 respectively. % TODO-nrf make sure this numbering is correct
+Similarly, $S_2 \sreduces \rstr{s_2}$ and $P_2 \treduces \str{s_2}$ for some $s_2 \in \lang{r_2}$.
+Therefore, $S \sreduces \rstr{s_1 s_2}$ by S-E-Concat and $\concat{P_1}{P_2} \treduces \str{s_1 s_2}$ by P-E-Concat.
+Finally, $s_1s_2 \in \lang{r_1}{r_2}$ by \ref{thm:regexprops}.
+
+Consider property (b).
+If $S_1 \err$ then $P_1 \err$ by induction, and it follows that $S \err$ and $P \err$ by respective error propagation rules.
+Similarly, if $S_2 \err$ then $P_2 \err$ and it follow that $S \err$ and $P \err$ by induction and propagation.
+
+\item[S-T-Replace:]
+Let $S = \rreplace{r}{S_1}{S_2}$ and suppose $\sctx \vdash S : \stringin{r'}$ for some $s$.
+By inversion of S-T-Replace, , $\sctx S_1 : \stringin{r_1}$ and $\sctx : \stringin{r_2}$ 
+such that $\lsubst{r}{r_1}{r_2} = r'$.
+By induction, there exists some $P_1, P_2$ such that $\trden{S_1} = P_1$, $\trden{S_2} = P_2$
+and either (a) or (b) holds.
+
+If (a) holds then $S_1 \sreduces \rstr{s_1}$ and $P_1 \treduces \str{s_1}$ for some $s_1 \in \lang{r_1}$,
+and similarly for $S_2, P_2$ and some $s_2 \in \lang{r_2}$.
+Therefore, by S-E-Replace, $S \sreduces \rstr{s}$ for some $s = \lreplace{r}{s_1}{s_2}$.
+Choose $P = \preplace{r}{s_1}{s_2}$.
+By a similar argument and P-E-Replace, $P \treduces \str{s}$ for some $s = \lreplace{r}{s_1}{s_2}$.
+What remains to be shown is $\lreplace{r}{s_1}{s_2} \in \lang{\lsubst{r}{r_1}{r_2}}$,
+which follows from Leamm D since $s_1 \in r_1$ and $s_2 \in r_2$.
+
+If (b) holds for $S_1$ and $P_1$, then $S \err$ and $P \err$ by propagation rules.
+Similarly, if (b) hols for $S_2$ and $P_2$ then $S \err$ and $P \err$ by propagation rules.
+
+\item[S-T-Coerce:]
+Let $S = \rcoerce{r}{S'}$ and suppose $\sctx \vdash \rcoerce{r}{S} : \stringin{r}$.
+By inverstion $\Psi \vdash S' : \stringin{r'}$ for an arbitrary $r'$.
+By induction there exists a $P'$ such that $\trden{S'} = P'$ and either (a) or (b)
+holds for $S'$ and $P'$.
+
+If (a) holds then $S' \sreduces \rstr{s'}$ and $P' \treduces \str{s'}$ for some $s' \in \lang{r'}$.
+Note that either $s' \in \lang{r}$ or $s' \not \in \lang{r}$ by property 2 of \ref{thm:regexprops}.
+Suppose $s' \in \lang{r}$. Then $\rcoerce{r}{S} \sreduces \rstr{S'}$ by S-E-Coerce.
+Choose $P = \check{\rx{r}}{P'}$ and note that $P \treduces \str{s'}$ by P-E-Coerce.
+Now suppose $s' \not \in \lang{r}$. Then $S \err$ and $P \err$ by P-E-Check-Err
+and S-E-Coerce-Err.
+
+Finally, if (b) holds then $S \err$ and $P \err$ by propagation.
 
@@
 \end{proof}
+
+% NRF commented out section 2 in order to determine the length of section 1.
+% Replace the pattern ^% with an emptty string in the selection starting at
+% the first empty line of section 2 and ending just before \end{document}.
 
@@
 
-\newcommand{\F}[1]{\textsf{#1}~}
-\newcommand{\FF}[1]{\textsf{#1}}
-\newcommand{\Q}{\FF{Arg}}
-\newcommand{\xlA}[1]{\lfloor #1 \rfloor_{\lambda}}
-\newcommand{\xA}[1]{$\lfloor #1 \rfloor_{\text{Ace}}$}
-\newcommand{\rtotau}[1]{\lfloor #1 \rfloor}
-
-\newcommand{\tremp}{{\tt tremp}}
-\newcommand{\trdot}{{\tt trdot}}
-\newcommand{\trchar}[1]{{\tt trchar}[#1]}
-\newcommand{\trseq}[2]{{\tt trseq}(#1; #2)}
-\newcommand{\tror}[2]{{\tt tror}(#1; #2)}
-\newcommand{\tlstr}[1]{{\tt tstr}[#1]}
-\begin{table*}[t]
-\centering
-\begin{tabular}{ l l l }
-$r$ & $\xlA{r}$ & \xA{r}\\
-\hline
-$\epsilon$ & $\tremp$ & \verb|""|\\
-$.$ & $\trdot$ & \verb|"."|\\
-$a$ & $\trchar{a}$ & \verb|"|$a$\verb|"|\\
-$r_1 \cdot r_2$ & $\trseq{\xlA{r_1}}{\xlA{r_2}}$ & \xA{r_1}\verb| + |\xA{r_2}\\
-$r_1 + r_2$ & $\tror{\xlA{r_1}}{\xlA{r_2}}$ & \verb|"(" + |\xA{r_1}\verb! + "|" + !\xA{r_2}\verb| + ")"|\\
-\\
-$\psi$ & $\xlA{\psi}$ & \xA{\psi}\\
-\hline
-${\tt string\_in}[r]$ & $\fvar{stringin}[\xlA{r}]$ & \verb|string_in[|\xA{r}\verb|]|\\
-\\
-  $S$ & $\xlA{S}$ & \xA{S} \\
-  \hline
-  ${\tt str\_in}[s]$ & $\FF{intro}[\FF{str}[s]]()$ & \verb|"|$s$\verb|"|\\
-  ${\tt concat}(S_1; S_2)$  & $\xlA{S_1}\cdot\FF{elim}[\tvar{concat}](\xlA{S_2})$ & \xA{S_1}\verb| + |\xA{S_2} \\
-  ${\tt subst}[r](S_1; S_2)$  & $\xlA{S_1}\cdot\FF{elim}[\tvar{subst}~\xlA{r}](\xlA{S_2})$ & \xA{S_1}\verb|.subst(|\xA{r}\verb|, |\xA{S_2}\verb|)|\\
-  ${\tt coerce}[r](S)$  & $\xlA{S}\cdot\FF{elim}[\tvar{coerce}~\xlA{r}]()$ & \xA{S}\verb|.coerce(|\xA{r}\verb|)|
-\end{tabular}
-\caption{Embeddings of the ${\tt string\_in}$ fragment into $\lamAce$ and Ace.}
-\end{table*}
-%
-\newcommand{\atjsynX}[3]{\Gamma \vdash_\fvalCtx #1 \Rightarrow #2 \leadsto #3}
-\newcommand{\atjanaX}[3]{\Gamma \vdash_\fvalCtx #1 \Leftarrow #2 \leadsto #3}
-\newcommand{\atjerrX}[1]{\Gamma \vdash_\fvalCtx #1~ \mathtt{error}}
-\begin{figure*}[t]
-\small
-$\fbox{\inferrule{}{\atjsynX{e}{\tau}{i}}}$~~~~
-$\fbox{\inferrule{}{\atjanaX{e}{\tau}{i}}}$~~~~
-%$\fbox{\inferrule{}{\atjerrX{e}}}$
-\begin{mathpar}
-\inferrule[att-flip]{
-	\atjsynX{e}{\tau}{i}
-}{
-	\atjanaX{e}{\tau}{i}
-}
-
-%\inferrule[att-var]{
-%	x \Rightarrow \tau \in \Gamma
+%\newcommand{\F}[1]{\textsf{#1}~}
+%\newcommand{\FF}[1]{\textsf{#1}}
+%\newcommand{\Q}{\FF{Arg}}
+%\newcommand{\xlA}[1]{\lfloor #1 \rfloor_{\lambda}}
+%\newcommand{\xA}[1]{$\lfloor #1 \rfloor_{\text{Ace}}$}
+%\newcommand{\rtotau}[1]{\lfloor #1 \rfloor}
+%
+%\newcommand{\tremp}{{\tt tremp}}
+%\newcommand{\trdot}{{\tt trdot}}
+%\newcommand{\trchar}[1]{{\tt trchar}[#1]}
+%\newcommand{\trseq}[2]{{\tt trseq}(#1; #2)}
+%\newcommand{\tror}[2]{{\tt tror}(#1; #2)}
+%\newcommand{\tlstr}[1]{{\tt tstr}[#1]}
+%\begin{table*}[t]
+%\centering
+%\begin{tabular}{ l l l }
+%$r$ & $\xlA{r}$ & \xA{r}\\
+%\hline
+%$\epsilon$ & $\tremp$ & \verb|""|\\
+%$.$ & $\trdot$ & \verb|"."|\\
+%$a$ & $\trchar{a}$ & \verb|"|$a$\verb|"|\\
+%$r_1 \cdot r_2$ & $\trseq{\xlA{r_1}}{\xlA{r_2}}$ & \xA{r_1}\verb| + |\xA{r_2}\\
+%$r_1 + r_2$ & $\tror{\xlA{r_1}}{\xlA{r_2}}$ & \verb|"(" + |\xA{r_1}\verb! + "|" + !\xA{r_2}\verb| + ")"|\\
+%\\
+%$\psi$ & $\xlA{\psi}$ & \xA{\psi}\\
+%\hline
+%${\tt string\_in}[r]$ & $\fvar{stringin}[\xlA{r}]$ & \verb|string_in[|\xA{r}\verb|]|\\
+%\\
+%  $S$ & $\xlA{S}$ & \xA{S} \\
+%  \hline
+%  ${\tt str\_in}[s]$ & $\FF{intro}[\FF{str}[s]]()$ & \verb|"|$s$\verb|"|\\
+%  ${\tt concat}(S_1; S_2)$  & $\xlA{S_1}\cdot\FF{elim}[\tvar{concat}](\xlA{S_2})$ & \xA{S_1}\verb| + |\xA{S_2} \\
+%  ${\tt subst}[r](S_1; S_2)$  & $\xlA{S_1}\cdot\FF{elim}[\tvar{subst}~\xlA{r}](\xlA{S_2})$ & \xA{S_1}\verb|.subst(|\xA{r}\verb|, |\xA{S_2}\verb|)|\\
+%  ${\tt coerce}[r](S)$  & $\xlA{S}\cdot\FF{elim}[\tvar{coerce}~\xlA{r}]()$ & \xA{S}\verb|.coerce(|\xA{r}\verb|)|
+%\end{tabular}
+%\caption{Embeddings of the ${\tt string\_in}$ fragment into $\lamAce$ and Ace.}
+%\end{table*}
+%%
+%\newcommand{\atjsynX}[3]{\Gamma \vdash_\fvalCtx #1 \Rightarrow #2 \leadsto #3}
+%\newcommand{\atjanaX}[3]{\Gamma \vdash_\fvalCtx #1 \Leftarrow #2 \leadsto #3}
+%\newcommand{\atjerrX}[1]{\Gamma \vdash_\fvalCtx #1~ \mathtt{error}}
+%\begin{figure*}[t]
+%\small
+%$\fbox{\inferrule{}{\atjsynX{e}{\tau}{i}}}$~~~~
+%$\fbox{\inferrule{}{\atjanaX{e}{\tau}{i}}}$~~~~
+%%$\fbox{\inferrule{}{\atjerrX{e}}}$
+%\begin{mathpar}
+%\inferrule[att-flip]{
+%	\atjsynX{e}{\tau}{i}
 %}{
-%	\atjsynX{x}{\tau}{x}
+%	\atjanaX{e}{\tau}{i}
 %}
 %
-%\inferrule[att-asc]{
-%    \tau \Downarrow_\fvalCtx \tau'\\
-%	\atjanaX{e}{\tau'}{i}
-%}{
-%	\atjsynX{e : \tau}{\tau'}{i}
-%}
-%
-%\inferrule[att-let-syn]{
-%	\atjsynX{e_1}{\tau_1}{i_1}\\
-%	\Gamma, x \Rightarrow \tau_1 \vdash_\fvalCtx e_2 \Rightarrow \tau_2 \leadsto i_2\\
-%	\trepof{\tau_1} \Downarrow_\fvalCtx \titype{\sigma_1}
-%}{
-%	\atjsynX{\F{let}x = e_1~\F{in}e_2}{\tau_2}{(\ilam{x}{\sigma_1}{i_2})~i_1}
-%}
-%
-%\inferrule[att-lam-ana]{
-%	\Gamma, x \Rightarrow \tau_1 \vdash_\fvalCtx e \Leftarrow \tau_2 \leadsto i\\
-%	\trepof{\tau_1} \Downarrow_\fvalCtx \titype{\sigma_1}
-%}{
-%	\atjanaX{\lambda x.e}{\ttype{arrow}{(\tau_1, \tau_2)}}{\ilam{x}{\sigma_1}{i}}
-%}
-%
-%\inferrule[att-lam]{
-%	\tau_1 \Downarrow_\fvalCtx \tau_1'\\
-%		\trepof{\tau_1'} \Downarrow_\fvalCtx \titype{\sigma}\\\\
-%	\Gamma, x \Rightarrow \tau_1' \vdash_\fvalCtx e \Rightarrow \tau_2 \leadsto i
-%%	i \hookrightarrow_\fvalCtx i'\\
-%%		\sigma \hookrightarrow_\fvalCtx \sigma'
-%%	\ddbar{\fvar{Arrow}}{\fvalCtx}{\trepof{\tau_1'}}{\sbar_1}\\
-%	%\delfromtau{$\Xi_0$}{\fvalCtx}{\tau_1'}{\sabs}\\\\
-%}{
-%	\atjsynX{\elam{x}{\tau_1}{e}}{\ttype{arrow}{(\tau_1', \tau_2)}}{\ilam{x}{\sigma'}{i}}
-%}
-%
-\inferrule[att-intro-ana]{
-	\vdash_\fvalCtx \FF{iana}(\fvar{tycon})=\taudef\\
-	\taudef~\taut{opidx}~\taut{tyidx}~((\Gamma; e_1)? :: \ldots :: (\Gamma; e_n)? :: []) \Downarrow_\fvalCtx \titerm{i}\\\\
-%	\trepof{\ttype{tycon}{\tauidx'}} \Downarrow_\fvalCtx \titype{\sigma}\\
-		\trepof{\ttype{tycon}{\taut{tyidx}}} \Downarrow_\fvalCtx \titype{\sigma}\\
-		\Gamma \vdash_\fvalCtx i : \sigma
-}{
-	\atjanaX{\FF{intro}[\taut{opidx}](e_1; \ldots; e_n)}{\ttype{tycon}{\taut{tyidx}}}{i}
-}
-
-%\inferrule[att-i-asc-ty]{
-%	\atjanaX{I}{\tau}{i}
-%}{
-%	\atjsynX{I : \tau}{\tau}{i}
-%}
-%
-%\inferrule[att-i-asc-tycon]{
-%	\vdash_\fvalCtx \FF{isyn}(\fvar{tycon})=\taudef\\\\
-%	\taudef~\tauidx~((\Gamma; e_1)? {::}{\ldots}{::}(\Gamma; e_n)? {::} []) \Downarrow_\fvalCtx \tden{\titerm{i}}{\ttype{tycon}{\tauidx'}}\\
-%			\trepof{\tau_1'} \Downarrow_\fvalCtx \titype{\sigma}\\
+%%\inferrule[att-var]{
+%%	x \Rightarrow \tau \in \Gamma
+%%}{
+%%	\atjsynX{x}{\tau}{x}
+%%}
+%%
+%%\inferrule[att-asc]{
+%%    \tau \Downarrow_\fvalCtx \tau'\\
+%%	\atjanaX{e}{\tau'}{i}
+%%}{
+%%	\atjsynX{e : \tau}{\tau'}{i}
+%%}
+%%
+%%\inferrule[att-let-syn]{
+%%	\atjsynX{e_1}{\tau_1}{i_1}\\
+%%	\Gamma, x \Rightarrow \tau_1 \vdash_\fvalCtx e_2 \Rightarrow \tau_2 \leadsto i_2\\
+%%	\trepof{\tau_1} \Downarrow_\fvalCtx \titype{\sigma_1}
+%%}{
+%%	\atjsynX{\F{let}x = e_1~\F{in}e_2}{\tau_2}{(\ilam{x}{\sigma_1}{i_2})~i_1}
+%%}
+%%
+%%\inferrule[att-lam-ana]{
+%%	\Gamma, x \Rightarrow \tau_1 \vdash_\fvalCtx e \Leftarrow \tau_2 \leadsto i\\
+%%	\trepof{\tau_1} \Downarrow_\fvalCtx \titype{\sigma_1}
+%%}{
+%%	\atjanaX{\lambda x.e}{\ttype{arrow}{(\tau_1, \tau_2)}}{\ilam{x}{\sigma_1}{i}}
+%%}
+%%
+%%\inferrule[att-lam]{
+%%	\tau_1 \Downarrow_\fvalCtx \tau_1'\\
+%%		\trepof{\tau_1'} \Downarrow_\fvalCtx \titype{\sigma}\\\\
+%%	\Gamma, x \Rightarrow \tau_1' \vdash_\fvalCtx e \Rightarrow \tau_2 \leadsto i
+%%%	i \hookrightarrow_\fvalCtx i'\\
+%%%		\sigma \hookrightarrow_\fvalCtx \sigma'
+%%%	\ddbar{\fvar{Arrow}}{\fvalCtx}{\trepof{\tau_1'}}{\sbar_1}\\
+%%	%\delfromtau{$\Xi_0$}{\fvalCtx}{\tau_1'}{\sabs}\\\\
+%%}{
+%%	\atjsynX{\elam{x}{\tau_1}{e}}{\ttype{arrow}{(\tau_1', \tau_2)}}{\ilam{x}{\sigma'}{i}}
+%%}
+%%
+%\inferrule[att-intro-ana]{
+%	\vdash_\fvalCtx \FF{iana}(\fvar{tycon})=\taudef\\
+%	\taudef~\taut{opidx}~\taut{tyidx}~((\Gamma; e_1)? :: \ldots :: (\Gamma; e_n)? :: []) \Downarrow_\fvalCtx \titerm{i}\\\\
+%%	\trepof{\ttype{tycon}{\tauidx'}} \Downarrow_\fvalCtx \titype{\sigma}\\
+%		\trepof{\ttype{tycon}{\taut{tyidx}}} \Downarrow_\fvalCtx \titype{\sigma}\\
 %		\Gamma \vdash_\fvalCtx i : \sigma
 %}{
-%	\atjsynX{\FF{intro}[\tauidx](e_1; \ldots; e_n)] :: \fvar{tycon}}{\ttype{tycon}{\tauidx'}}{i'}
+%	\atjanaX{\FF{intro}[\taut{opidx}](e_1; \ldots; e_n)}{\ttype{tycon}{\taut{tyidx}}}{i}
 %}
 %
-\inferrule[att-elim-syn]{
-	\atjsynX{e}{\ttype{tycon}{\taut{tyidx}}}{i}\\
-	\vdash_\fvalCtx \FF{esyn}(\fvar{tycon})=\taudef\\\\
-	\taudef~\taut{opidx}~\taut{tyidx}~\titerm{i}~((\Gamma; e)? :: (\Gamma; e_1)? :: \ldots :: (\Gamma; e_n)? :: []) \Downarrow_\fvalCtx (\tau, \titerm{i'})\\\\
-			\trepof{\tau} \Downarrow_\fvalCtx \titype{\sigma}\\
-	\Gamma \vdash_\fvalCtx i' : \sigma
-}{
-	\atjsynX{e\cdot\FF{elim}[\taut{opidx}](e_1; \ldots; e_n)}{\tau}{i'}
-}
-\end{mathpar}
-%\vspace{-10px}
-\caption{The bidirectional active typechecking and translation judgements.}
-\label{atj}
-\end{figure*}
-\newcommand{\tlevalX}[2]{#1 \Downarrow_\fvalCtx #2}
-\begin{figure*}[t]
-\small
-$\fbox{\inferrule{}{\tau \Downarrow_\fvalCtx \tau'}}$~~~~
-\begin{mathpar}
-\inferrule[trstr-eval]{ }{\tlevalX{\tlstr{s}}{\tlstr{s}}}
-
-\inferrule[tremp-eval]{ }{\tremp \Downarrow_\fvalCtx \tremp}
-
-\cdots
-
-\inferrule[tror-eval]{
-	\tlevalX{\tau_1}{\tau_1'}\\
-	\tlevalX{\tau_2}{\tau_2'}
-}{
-	\tlevalX{\tror{\tau_1}{\tau_2}}{\tror{\tau_1'}{\tau_2'}}
-}
-
-trmatch
-
-trlsubst
-
-trsublang
-
-%\inferrule[repof]{
-%	\tau \Downarrow_\fvalCtx \ttype{tycon}{\tauidx}\\
-%	\vdash_\fvalCtx \FF{rep}(\fvar{tycon}) = \taurep\\
-%	\taurep~\tauidx \Downarrow_\fvalCtx \titype{\sigma}
+%%\inferrule[att-i-asc-ty]{
+%%	\atjanaX{I}{\tau}{i}
+%%}{
+%%	\atjsynX{I : \tau}{\tau}{i}
+%%}
+%%
+%%\inferrule[att-i-asc-tycon]{
+%%	\vdash_\fvalCtx \FF{isyn}(\fvar{tycon})=\taudef\\\\
+%%	\taudef~\tauidx~((\Gamma; e_1)? {::}{\ldots}{::}(\Gamma; e_n)? {::} []) \Downarrow_\fvalCtx \tden{\titerm{i}}{\ttype{tycon}{\tauidx'}}\\
+%%			\trepof{\tau_1'} \Downarrow_\fvalCtx \titype{\sigma}\\
+%%	\Gamma \vdash_\fvalCtx i : \sigma
+%%}{
+%%	\atjsynX{\FF{intro}[\tauidx](e_1; \ldots; e_n)] :: \fvar{tycon}}{\ttype{tycon}{\tauidx'}}{i'}
+%%}
+%%
+%\inferrule[att-elim-syn]{
+%	\atjsynX{e}{\ttype{tycon}{\taut{tyidx}}}{i}\\
+%	\vdash_\fvalCtx \FF{esyn}(\fvar{tycon})=\taudef\\\\
+%	\taudef~\taut{opidx}~\taut{tyidx}~\titerm{i}~((\Gamma; e)? :: (\Gamma; e_1)? :: \ldots :: (\Gamma; e_n)? :: []) \Downarrow_\fvalCtx (\tau, \titerm{i'})\\\\
+%			\trepof{\tau} \Downarrow_\fvalCtx \titype{\sigma}\\
+%	\Gamma \vdash_\fvalCtx i' : \sigma
 %}{
-%	\FF{repof}(\tau) \Downarrow_\fvalCtx \titype{\sigma}
+%	\atjsynX{e\cdot\FF{elim}[\taut{opidx}](e_1; \ldots; e_n)}{\tau}{i'}
 %}
-%
-\inferrule[syn]{
-    \tau \Downarrow_\fvalCtx (\Gamma; e)?\\
-	\atjsynX{e}{\tau}{\iota}\\\\
-    [\tau/\tvar{t}_{ty}, \titerm{\iota}/\tvar{t}_{trans}]\tau_2 \Downarrow_\fvalCtx \tau_2'
-}{
-	\FF{syn}(\tau_1; \tvar{t}_{ty}, \tvar{t}_{trans}.\tau_2) \Downarrow_\fvalCtx \tau_2'
-}
-
-\inferrule[ana]{
-	\tau_1 \Downarrow_\fvalCtx (\Gamma; e)?\\
-	\tau_2 \Downarrow_\fvalCtx \tau_2'\\
-	\atjanaX{e}{\tau_2'}{\iota}\\\\
-	[\titerm{\iota}/\tvar{t}_{trans}]\tau_3 \Downarrow_\fvalCtx \tau_3'
-}{
-	\FF{ana}(\tau_1; \tau_2; \tvar{t}_{trans}.\tau_3) \Downarrow_\fvalCtx \tau_3'
-}
-
-TODO: add rx, str stuff
-\end{mathpar}
-\caption{\small Normalization semantics for the type-level language.
+%\end{mathpar}
+%%\vspace{-10px}
+%\caption{The bidirectional active typechecking and translation judgements.}
+%\label{atj}
+%\end{figure*}
+%\newcommand{\tlevalX}[2]{#1 \Downarrow_\fvalCtx #2}
+%\begin{figure*}[t]
+%\small
+%$\fbox{\inferrule{}{\tau \Downarrow_\fvalCtx \tau'}}$~~~~
+%\begin{mathpar}
+%\inferrule[trstr-eval]{ }{\tlevalX{\tlstr{s}}{\tlstr{s}}}
+%
+%\inferrule[tremp-eval]{ }{\tremp \Downarrow_\fvalCtx \tremp}
+%
+%\cdots
+%
+%\inferrule[tror-eval]{
+%	\tlevalX{\tau_1}{\tau_1'}\\
+%	\tlevalX{\tau_2}{\tau_2'}
+%}{
+%	\tlevalX{\tror{\tau_1}{\tau_2}}{\tror{\tau_1'}{\tau_2'}}
+%}
+%
+%trmatch
+%
+%trlsubst
+%
+%trsublang
+%
+%
+%%\inferrule[repof]{
+%%	\tau \Downarrow_\fvalCtx \ttype{tycon}{\tauidx}\\
+%%	\vdash_\fvalCtx \FF{rep}(\fvar{tycon}) = \taurep\\
+%%	\taurep~\tauidx \Downarrow_\fvalCtx \titype{\sigma}
+%%}{
+%%	\FF{repof}(\tau) \Downarrow_\fvalCtx \titype{\sigma}
+%%}
+%%
+%\inferrule[syn]{
+%    \tau \Downarrow_\fvalCtx (\Gamma; e)?\\
+%	\atjsynX{e}{\tau}{\iota}\\\\
+%    [\tau/\tvar{t}_{ty}, \titerm{\iota}/\tvar{t}_{trans}]\tau_2 \Downarrow_\fvalCtx \tau_2'
+%}{
+%	\FF{syn}(\tau_1; \tvar{t}_{ty}, \tvar{t}_{trans}.\tau_2) \Downarrow_\fvalCtx \tau_2'
+%}
+%
+%\inferrule[ana]{
+%	\tau_1 \Downarrow_\fvalCtx (\Gamma; e)?\\
+%	\tau_2 \Downarrow_\fvalCtx \tau_2'\\
+%	\atjanaX{e}{\tau_2'}{\iota}\\\\
+%	[\titerm{\iota}/\tvar{t}_{trans}]\tau_3 \Downarrow_\fvalCtx \tau_3'
+%}{
+%	\FF{ana}(\tau_1; \tau_2; \tvar{t}_{trans}.\tau_3) \Downarrow_\fvalCtx \tau_3'
+%}
+%
+%TODO: add rx, str stuff
+%\end{mathpar}
+%\caption{\small Normalization semantics for the type-level language.
 Missing rules (including error propagation rules and normalization of quoted internal terms and types) are unsurprising and will be given later.}
-\label{tleval}
-\end{figure*}
-\begin{figure}[t]
-\small\begin{flalign}
-& \F{tycon}\fvar{stringin}~\F{of}\FF{R}~\{\\
-& \quad \F{iana}\{\tlam{opidx}{\FF{String}}{
-	\tlam{tyidx}{\FF{R}}{
-	\tlam{a}{\klist{\Q}}{\\
-& \quad\quad \tvar{arity0}~\tvar{a}~(\tvar{check}~\tvar{opidx}~\tvar{tyidx}~\titerm{\iup{\tvar{opidx}}})
-	}}
-}\}\\
-& \quad \F{esyn}\{\tlam{opidx}{\kunit+(\FF{R}+\FF{R})}{
-	\tlam{a}{\klist{\Q}}{\\
-& \quad\quad \tsumcase{\tvar{opidx}}{\_}{\\
-& \quad\quad\quad \tvar{arity2}~\tvar{a}~\tlam{a1}{\Q}{\tlam{a2}{\Q}{\\
-& \quad\quad\quad\quad \tvar{rsyn}~\tvar{a1}~\tlam{r1}{\FF{R}}{\tlam{i1}{\kITerm}{~}}\\
-& \quad\quad\quad\quad \tvar{rsyn}~\tvar{a2}~\tlam{r2}{\FF{R}}{\tlam{i2}{\kITerm}{~}}\\
-& \quad\quad\quad\quad\quad (\ttype{stringin}{\FF{rseq}(\tvar{r1}; \tvar{r2})}, \\
-& \quad\quad\quad\quad\quad\titerm{{\tt iconcat}(\iup{\tvar{i1}}; \iup{\tvar{i2}})})
-}}\\
-& \quad\quad}{opidx'}{d}}
-}\}
-\};\\
-& \F{def}\tvar{concat} = \tinl{\kunit, \FF{R}+\FF{R}}{\tunit};\\
-& \F{def}\tvar{subst} = \tlam{r}{\FF{R}}{\tinr{\kunit, \FF{R}+\FF{R}}{\tinl{\FF{R},\FF{R}}{\tvar{r}}}};\\
-& \F{def}\tvar{coerce} = \tlam{r}{\FF{R}}{\tinr{\kunit, \FF{R}+\FF{R}}{\tinr{\FF{R},\FF{R}}{\tvar{r}}}}
-\end{flalign}
-\caption{Definition of $\phi_S$, which enables the embedding of fragment $S$ into $\lamAce$.}
-\end{figure}
-%\begin{figure}
-%\[
-%\begin{array}{lcl}
-%% & & Definition & Kind\\
-%\tvar{concat} & := & \tinl{\kunit, \FF{R}+\FF{R}}{\tunit}\\%& \kunit + (R + R)\\
-%\tvar{replace} & := & \tlam{r}{\FF{R}}{\tinr{\kunit, \FF{R}+\FF{R}}{\tinl{\FF{R},\FF{R}}{\tvar{r}}}}\\% & \karrow{a}{b}\\
-%\tvar{coerce} & := & \tlam{r}{\FF{R}}{\tinr{\kunit, \FF{R}+\FF{R}}{\tinr{\FF{R},\FF{R}}{\tvar{r}}}}\\
-%\end{array}
-%\]
-%\caption{Definitions}
+%\label{tleval}
+%\end{figure*}
+%\begin{figure}[t]
+%\small\begin{flalign}
+%& \F{tycon}\fvar{stringin}~\F{of}\FF{R}~\{\\
+%& \quad \F{iana}\{\tlam{opidx}{\FF{String}}{
+%	\tlam{tyidx}{\FF{R}}{
+%	\tlam{a}{\klist{\Q}}{\\
+%& \quad\quad \tvar{arity0}~\tvar{a}~(\tvar{check}~\tvar{opidx}~\tvar{tyidx}~\titerm{\iup{\tvar{opidx}}})
+%	}}
+%}\}\\
+%& \quad \F{esyn}\{\tlam{opidx}{\kunit+(\FF{R}+\FF{R})}{
+%	\tlam{a}{\klist{\Q}}{\\
+%& \quad\quad \tsumcase{\tvar{opidx}}{\_}{\\
+%& \quad\quad\quad \tvar{arity2}~\tvar{a}~\tlam{a1}{\Q}{\tlam{a2}{\Q}{\\
+%& \quad\quad\quad\quad \tvar{rsyn}~\tvar{a1}~\tlam{r1}{\FF{R}}{\tlam{i1}{\kITerm}{~}}\\
+%& \quad\quad\quad\quad \tvar{rsyn}~\tvar{a2}~\tlam{r2}{\FF{R}}{\tlam{i2}{\kITerm}{~}}\\
+%& \quad\quad\quad\quad\quad (\ttype{stringin}{\FF{rseq}(\tvar{r1}; \tvar{r2})}, \\
+%& \quad\quad\quad\quad\quad\titerm{{\tt iconcat}(\iup{\tvar{i1}}; \iup{\tvar{i2}})})
+%}}\\
+%& \quad\quad}{opidx'}{d}}
+%}\}
+%\};\\
+%& \F{def}\tvar{concat} = \tinl{\kunit, \FF{R}+\FF{R}}{\tunit};\\
+%& \F{def}\tvar{subst} = \tlam{r}{\FF{R}}{\tinr{\kunit, \FF{R}+\FF{R}}{\tinl{\FF{R},\FF{R}}{\tvar{r}}}};\\
+%& \F{def}\tvar{coerce} = \tlam{r}{\FF{R}}{\tinr{\kunit, \FF{R}+\FF{R}}{\tinr{\FF{R},\FF{R}}{\tvar{r}}}}
+%\end{flalign}
+%\caption{Definition of $\phi_S$, which enables the embedding of fragment $S$ into $\lamAce$.}
 %\end{figure}
-
-\subsection{Background: Ace}
-TODO: Make a TR out of the OOPSLA submission.
-\subsection{Explicit Conversions}
-\subsection{Adding Subtyping to Ace}
-\subsection{Theory}
-
-\section{Related Work}
-
-\section{Discussion}
-
-\bibliographystyle{abbrv}
-
-% The bibliography should be embedded for final submission.
-
-\bibliography{../research}
-%\softraggedright
-%P. Q. Smith, and X. Y. Jones. ...reference text...
+%%\begin{figure}
+%%\[
+%%\begin{array}{lcl}
+%%% & & Definition & Kind\\
+%%\tvar{concat} & := & \tinl{\kunit, \FF{R}+\FF{R}}{\tunit}\\%& \kunit + (R + R)\\
+%%\tvar{replace} & := & \tlam{r}{\FF{R}}{\tinr{\kunit, \FF{R}+\FF{R}}{\tinl{\FF{R},\FF{R}}{\tvar{r}}}}\\% & \karrow{a}{b}\\
+%%\tvar{coerce} & := & \tlam{r}{\FF{R}}{\tinr{\kunit, \FF{R}+\FF{R}}{\tinr{\FF{R},\FF{R}}{\tvar{r}}}}\\
+%%\end{array}
+%%\]
+%%\caption{Definitions}
+%%\end{figure}
+%
+%\subsection{Background: Ace}
+%TODO: Make a TR out of the OOPSLA submission.
+%\subsection{Explicit Conversions}
+%\subsection{Adding Subtyping to Ace}
+%\subsection{Theory}
+%
+%\section{Related Work}
+%
+%\section{Discussion}
+%
+%
+%\bibliographystyle{abbrv}
+%
+%% The bibliography should be embedded for final submission.
+%
+%\bibliography{../research}
+%%\softraggedright
+%%P. Q. Smith, and X. Y. Jones. ...reference text...
+%
+%
 

### 2014-08-04 sanitation-psp14/sanitation-psp14.tex
Ready for comments on first section & intro, mod citations.

@@
 
+\newcommand{\iso}{\cong}
 %%
@@
 \section{Introduction}\label{intro}
-In web applications and other settings, incorrect input sanitation often causes
-security vulnerabilities. In fact, ...OWASP says it's important... .
-Mention CVE stats.
+Improper input sanitation is a leading cause of security vulnerabilities in web
+applications [OWASP].
 For this reason, modern web frameworks and libraries use various techniques to
-ensure proper sanitation of arbitrary user input. On the rare occasions when
-these methods are unavailable or insufficient, developers (hopefully) create 
+ensure proper sanitation of arbitrary user input. When
+these methods are unavailable or insufficient, developers create 
 ad hoc sanitation algorithms. Im most cases, sanitation algorithms -- ad hoc 
@@
 
+%%
+% This is the paragraph which ties the paper back to the CFP.
+%
+% For refernce, here's a copy of the cfp:
+%The importance of security and privacy in software engineering is now a pressing concern. 
+%Over the last decade, many programmers have recognized the importance of including 
+%security and privacy requirements at project start. This has led to efforts 
+%in secure/security engineering and privacy engineering, which focus on guidelines and 
+%best practices that can be used at the design stage to create safer code. 
+%Unfortunately, these disciplines are not pervasive and still in their infancy.
+%The goal of this workshop is to make security and privacy first class citizens in programming and programming languages.
+
+%This forum aims to gather research and industry professionals to further the discussion on how to:
+%   codify the principles from secure and privacy engineering into programming language constructs and or tools, 
+%   create programming languages that have security and privacy as foundational tenets, and 
+%   create/codify constructs or tools that enables secure and privacy-preserving (business) operations.
+%   All submissions connected to the goals above, and their associated topics, are encouraged.
+%%
+
 This paper presents a language extension, definable in the Ace programming language,
 for ensuring that input sanitation algorithms are implemented correctly with
-respect to use site specifications. If use site specifications are sufficient, 
-then type checking ensures the absence of vulnerabilities and other bugs which
-arise from improper sanitation. Our extension focuses on sanitation algorithms 
-which aim to prevent command-injection style attacks (e.g. SQL injection or RFI).
+respect to use site specifications.
+
+\emph{Instead of developing a top-down and holistic solution to the always-shifting security challenge,
+we propose an approach which allows developers to incorporate light-weight static
+analyses which best validate their privacy or security requirements.}
+In this paper, we present a type system which ensures the absence of vulnerabilities and other bugs
+which arise from imporper input sanitation.
+We then demonstrate how this system is embeddable in an extensible type system.
 
@@
 which captures idiomatic sanitation algorithms. Type safety in our system relies upon 
-several closure and decidability results about regular languages.
+several closure and decidability results about operations on 
+regular languages which correspond to typical operations for sanitizing strings.
 
 Libraries and frameworks available in functional programming language communities often
-make claims about security and sometimes even sophistically mention sophisticated
-type systems as evidence of freedom from injection-style attacked.
-However, even the specification that input is always sanitized properly before use is not
-actually captured by the type system or anywhere else; in fact, the full specification of these
-algorithms is rarely characterized by anything more specific than the type
-String $\rightarrow$ String, which is a class of safe input sanitation algorithms
-only in the most degenerate case.
-
-A number of research languages provide actually static guarantees that a program
+make claims about security and sometimes even mention sophisticated
+type systems as evidence of freedom from injection-style attacks.
+In reality, many of these approaches ultimately depend upon the correct implementation
+of library or framework input sanitation functions. In these cases, our type system
+provides a stronger (compile-time) guarantee and is additionally orthogonal because
+the correctness of the library or framework itself could be checked by implementing its
+sanitation functions in terms of our system.
+
+A number of research languages provide static guarantees that a program
 is free of input sanitation vulnerabilities. Most rely on some form of information
-flow. TODO-nrf  give citations and examples. Our extension to Ace differs from
+flow TODO-nrf citations. Our extension to Ace differs from
 these systems in the ways following:
@@
   \item Our system is defined as a library in terms of an extensible type system,
-  as opposed to a stand-alone language. This is important for two reasons.
-  First, our system is one of the first large examples written in Ace, and serves
-  as an extended case study. Second, although our approach requires
-  developers to adopt a new language, it does not require developers to adopt
-  a language specifically for web development.
+  as opposed to a stand-alone language. Instead of introducing new technologies
+  and methodologies for addressing security problems, we provide a light-weight
+  static analysis \emph{based upon existing approaches to the problem} which are
+  already well-understood by developers. We believe extensibility is a viable
+  and attractive alternative to highly specialized programming languages.
   \item Ace is implemented in Python and shares its grammar. Since Python is a 
@@
   and adopted technologies is far lower than is e.g. Ur/Web's.
-  \item In general, our is a composable, light-weight and
-  natural solution to a single problem -- rather than a comprehensive 
-  solution to the web security problem. Our goal is to demonstrate that extensible
-  type systems can capture static properties of common idioms, and thereby ensure
-  safety without introducing much additional complexity.
 \end{itemize}
 
-Finally, incorporating regular expressions into the type system is not novel.
+Incorporating regular expressions into the type system is not novel.
 The XDuce system \cite{pierce} typechecks XML schemas using regular expressions.
@@
 First, our system  is defined within an extensible type system;
-this first difference introduced an interesting class of design problems (see \S ???).
-TODO-nrf this is the second time this is used as a warrant. Factor out this argument
-and place it at the top???
-Second, our focus on security required novel design decisions; for instance, the
-filter elimination form is unique to Ace.
+second, and more importantly, we have demonstrated that regular expression types
+are applicable to the web security domain.
 
@@
 
-TODO-nrf real outline!
 \begin{itemize}
-  \item In \S 2, we define the type system's static and dynamic semantics.
-  \item Section 3 recalls some classical results about regular expressions and presents meta-theory for our system, including
-    the main soundness theorem for $\lcs$.
-  \item Finally, \S 4 discusses our implemention of $\lcs$ as a type system extension  within the Ace programming language.
+  \item In \S 2, we define the type system which is embedded in Ace. We include a type safety proof for the string
+    segment of this language and prove the correctness of a translation to an underlying langguage $P$. In our theory,
+    $P$ is a simply typed lambda calculus equipped with a minimal regular expression library; in an implementation, $P$
+    stands in for Python or another underlying general-purpose programming language.
+  \item In \S 3, we discuss our implemention of this translation as a type system extension  within the Ace programming language.
 \end{itemize}
@@
 
-The $\lcs$ language is characterized by a type of strings indexed by regular
+The $\lambdas$ language is characterized by a type of strings indexed by regular
 expressions, together with operations on such strings which correspond to common
 input sanitation patterns.
-
-This section presents the grammar and semantics of $\lcs$.
-The semantics are defined in terms of an internal language with at least strings and a regex filter function.
-These constraints are captured by the internal term valuations.
-The internal language does not necessarily need a regex filter function because
-any dynamic conversion is easily definable using a combination of filters and safe
-casting.
-
-Note that we never insert an internal check where the type of the string implies
-that a check must succeed. Furthermore, expensive calculations (such as language
-inclusion or ${\tt dsubst}$, all occur at compile time.
-Since sanitation problems are generally expressible with smalle, simple expressions,
-we believe that the compile-time overhead is not significant enough to prohibiour target usecase. Informal experimentation with our implementation seems to
-support this assertion.
-
-Due to Ei-Replace, all of our meta-theory depends upon a correct implementation
-of both regular expression replacement and inclusion checking in the internal language.
-We believe this assumption is okay for two reasons. First, our system is still
-superior to the status quo, which relies upon the correctness of these libraries
-\emph{in addition to} correct application logic. Second, regular expression libraries are
-generally well-tested and there exist verified implementations.
+This section presents the grammar, typing rules and operational semantics for
+$\lambdas$ as well as an underlying language $\lambdap$.
+
+The system $\lambdas$ \ref{fig:lambdas} is the simply typed lambda calculus extended with \emph{regular expression types}, which are string
+types ensuring a string belongs to a specified language. For instance, 
+$S : \rstr{r}$ reads ``$s$ is a string matching $r$".
+the system includes an operation for replacing all instances of a pattern $r$ in a string $s_1$ with another string $s_2$. 
+Input sanitation algorithms -- as implemented by developers or within popular libraries and frameworks -- are often implemented in terms of this replace operation.
+For instance, a developer might all potentially unsafe charapcters with excaped versions of the same character.
+Regular expression types are used both to specify input sanitation algorithms, and at use sites as specifications.
+Note that runtime error states ($S \err$) are introduced by coercion, not by replacement.
+
+The language $\lambdap$ \ref{fig:lambdap} is a simple functional language extended with a minimal regular expression library.
+Any general purpose programming language could stand in for $\lambdap$; for instance, SML has a regular expression library.
+In an implementation, our correctness results are modulo the underlying language's correct implementation of regular expression matching (see P-E-Replace).
+
+Finally, we define a translation from our type system $\lambdas$ into $\lambdap$.
+
+%Note that we never insert an internal check where the type of the string implies
+%that a check must succeed. Furthermore, expensive calculations (such as language
+%inclusion or ${\sf rreplace}$) all occur at compile time.
+%Since sanitation problems are generally expressible with smalle, simple expressions,
+%we believe that the compile-time overhead is not significant enough to prohibiour target usecase. Informal experimentation with our implementation seems to
+%support this assertion.
 
@@
 
-there's extra space here because of the next page...
-
 %The $\lcs$ language gives static semantics for common regular expression library
@@
 \addtocounter{figure}{4}
-\begin{figure*}
+\begin{figure*}\label{fig:trans}
 
@@
 
-\begin{figure}[t]
+\begin{figure}[t]\label{fig:lambdas}
 \small
@@
 
-\begin{figure}[t]
+\begin{figure}[t]\label{fig:lambdap}
 \small
@@
 \item[(2):] For all strings $s$ and expressions $r$, either $s \in \lang{r}$ or $s \not \in \lang{r}$.
-\end{description}
 \item[(3):] Regular languages are closed under complements and concatenation.
 \item[(4):] The regular expressions correspond bijectively to the regular languages.
+\end{description}
 \end{lem}
@@
 \begin{proof}
-TODO-nrf 
+By induction on $r$ and closure properties of regular expressions.
 \end{proof}
@@
 \begin{proof}
-TODO-nrf 
+The proof proceeds by structural induction on $r$.
+
+\begin{description}
+\item[case $r=\alpha$.]
+If $s_1 = \alpha$ then $\alpha \in \lang{r_1}$ by assumption.
+Therefore, $\lsubst{r}{s_1}{s_2} = \lsubst{\alpha}{\alpha}{s_2} = s_2$ and
+$\lreplace{r}{r_1}{r_2} = \lreplace{\alpha}{r_1}{r_2}$. 
+Since $s_1 = \alpha$ and $\alpha \in \lang{r_1}$, $r_1 \iso \alpha | r_1'$ for some $r_1'$.
+Therefore, $\lreplace{\alpha}{r_1}{r_2} \iso \lreplace{\alpha}{\alpha|r_1'}{r_2}$ by Lemma X(1).
+Finally, $s_2 \in \lang{r_2}$ which implies $s_1 \in \lang{r_2|r'}$.
+If $s_1 \not = \alpha$ the $\lsubst{r}{s_1}{s_2} = s_1$ and $\lreplace{\alpha}{r_1}{r_2} = r_1$.
+
+\item[case $r = a|b$].
+Note that $\subst{a|b}{s_1}{s_2} = \lsubst{a}{\lsubst{b}{s_1}{s_2}}{s_2}$
+and $\lsubst{b}{s_1}{s_2} \in \lreplace{b}{r_1}{r_2}$ by induction.
+Therefore, $\lsubst{a|b}{s_1}{s_2}$ $\in \lreplace{a}{\lreplace{b}{r_1}{r_2}}{r_2}$ by induction and the definition of lreplace.
+Finally, applying definitions once more, $\lsubst{a|b}{s_1}{s_2} \in \lreplace{a|b}{r_1}{r_2}$.
+
+\item[case $r = ab$].
+By a similar argument to the disjunctive case.
+
+\item[case $r = a*$].
+By considering the once unwinding of $a*$, noting that $s_1$ and $s_2$ are finite.
+\end{description}
 \end{proof}
@@
   Let $S$ be a term in the source language. If $\sctx \vdash S : \stringin{r}$ 
-  then $S \sreduces \rstr{s}$ and $\rstr{s} : \stringin{r}$, or else $S \sreduces \err$.
+  then $S \sreduces \rstr{s}$ and $\rstr{s} : \stringin{r}$, or else $S \err$.
 \end{thm}
@@
 By inversion, $\sctx \vdash S_1 : \stringin{r_1}$ and $\sctx \vdash S_1 : \stringin{r_2}$.
-It follows by induction that either $S_1 \sreduces \err$, $S_2 \sreduces \err$, 
+It follows by induction that either $S_1 \err$, $S_2 \err$, 
 or $S_1 \sreduces \rstr{s_1}$ and $S_2 \sreduces \rstr{s_2}$ for some $s_1 \in \lang{r_1}$ and $s_2 \in \lang{r_2}$.
 In the latter case $S \reduces \rstr{s_1s_2}$ by S-E-Concat and $\sctx \vdash \rstr{s_1s_2} : \tstr{r_1r_2}$ by \ref{thm:regexprops}.
-In the former cases, $S \sreduces \err$.
+In the former cases, $S \err$.
 
@@
 By inversion $\sctx \vdash S_1 : \stringin{r_1}$ and $\sctx \vdash S_2 : \stringin{r_2}$ such that $\lsubst{r}{r_1}{r_2} = r'$.
-By induction, $S_1 \sreduces \err$, $S_2 \sreduces \err$ 
+By induction, $S_1 \err$, $S_2 \err$ 
 or $S_1 \sreduces \rstr{s_1}$ and $S_2 \sreduces \rstr{s_2}$ such that 
@@
 So by S-T-String-I, $\rstr{s} : \stringin{r'}$.
-In the former cases, $S \sreduces \err$.
+In the former cases, $S \err$.
 
@@
 By inversion, $\sctx \vdash S_1 : \stringin{r'}$.
-By induction, $S_1 \sreduces \err$ or $S_1 \sreduces \rstr{s}$.
-In the former case $S \sreduces \err$ by propagation rules.
+By induction, $S_1 \err$ or $S_1 \sreduces \rstr{s}$.
+In the former case $S \err$ by propagation rules.
 In the latter case we have by property 2 of \ref{thm:regexprops} that $s \in \lang{r}$ or else
@@
 If $s \in \lang{r}$ then $\rstr{s} : \stringin{r}$.
-If $s \not \in \lang{r}$ then $S \sreduces \err$. 
+If $s \not \in \lang{r}$ then $S \err$. 
 \end{description}
@@
 % the first empty line of section 2 and ending just before \end{document}.
+
+Papers that needs to be cited in this section:
+\begin{itemize}
+  \item Ur/Web OSDI paper
+  \item Jif?
+  \item OWASP
+  \item XDuce and related papers. 
+  \item src?
+  \item Ace or Wyvern paper?
+  \item hotsos?
+  \item Haskell extension paper
+  \item Maybe some popular FOSS libraries/frameworks that do input sanitation?
+\end{itemize}
 

### 2014-08-05 sanitation-psp14/sanitation-psp14.tex
Modified related work and intro section to de-emphasize section 3.

@@
 \section{Introduction}\label{intro}
-Improper input sanitation is a leading cause of security vulnerabilities in web
-applications [OWASP].
-For this reason, modern web frameworks and libraries use various techniques to
-ensure proper sanitation of arbitrary user input. When
-these methods are unavailable or insufficient, developers create 
-ad hoc sanitation algorithms. Im most cases, sanitation algorithms -- ad hoc 
-or otherwise -- are ultimately implemented using the language's regular
-expression capabilities. Therefore, a system capable of statically 
-checking properties about operations performed using regular expressions will
-be expressive enough to capture real-world implementations of sanitation 
-algorithms.
-
-Input sanitation is the problem of ensuring that an arbitrary string
-is coerced into a safe form before potentially unsafe use. For example, 
-preventing SQL injection attacks requires ensuring that any string
-coming from user input to a query does not contain unescaped SQL. 
-The point at which arbitrary user input is concatenated into a SQL query is called a use site.
-Although we believe are general approach extends to a wider class of problems
-(e.g. sanitation algorithms for preventing XSS attacks might be definable using
-regular tree languages), these generalizations are beyond the scope of the 
-present paper.
+Improper input sanitation is a leading cause of security vulnerabilities in web applications [OWASP].
+Command injection attacks exploit improper input sanitation by inserting malicious code into an otherwise benign command.
+Modern web frameworks, libraries, and database abstraction layers attempt to ensure proper sanitation of user input.
+When these methods are unavailable or insufficient, developers implement custom sanitation techniques.
+Im both cases, sanitation algorithms are implemented using the language's regular expression capabilities and usually \emph{replace} potentially unsafe strings with equivalent escaped strings.
+
+In this paper, we present a type system for implementing and statically checking input sanitation techniques.
+Our solution suggests a more general approach to the integration of security concerns into programming language design.
+This approach is characterized by \emph{composable} type system extensions which \emph{complement} existing and well-understood solutions with compile-time checks.
+
+To demonstrate this approach, we present a simply typed lambda calculus with \emph{constrained strings}; that is, a set of string types parameterized by regular expressions.
+If $s : \stringin{r}$, then $s$ is a string matching the language $r$.
+Additionally, we include an operation $\rreplace{r}{s_1}{s_2}$ which corresponds to the replace mechanism available in most regular expression libraries; that is, any substring of $s_1$ matching $r$ is replaced with $s_2$.
+The type of this expression is the computed, and is likely ``smaller" or more constrained than the type of $s_1$.
+Libraries, frameworks or functions which construct and execute commands containing input can specify a safe subset $\stringin{r_{\sf spec}}$ of strings, and input sanitation algorithms can construct such a string using ${\sf rreplace}$ or, optionally, by coercion (in which case a runtime check is inserted).
+We also show how this simple can be translated into a host language containing a regular expression library such that the safety guarantee of the extended language is preserved.
+
+Summarily, we present a simple type system extension which ensures the absence of input sanitation vulnerabilities by statically checking input sanitation algorithms which use an underlying regular expression library.
+This approach is \emph{composable} in the sense that it is a conservative extension.
+This approach is also \emph{complementary} to existing input sanitation techniques which use string replacement for input sanitation.
 
@@
 
-This paper presents a language extension, definable in the Ace programming language,
-for ensuring that input sanitation algorithms are implemented correctly with
-respect to use site specifications.
-
-\emph{Instead of developing a top-down and holistic solution to the always-shifting security challenge,
-we propose an approach which allows developers to incorporate light-weight static
-analyses which best validate their privacy or security requirements.}
-In this paper, we present a type system which ensures the absence of vulnerabilities and other bugs
-which arise from imporper input sanitation.
-We then demonstrate how this system is embeddable in an extensible type system.
-
 \subsection{Related Work and Alternative Approaches}
 
-The input sanitation problem is well-understood. There exist a large number
-of techniques and technologies, proposed by both practitioners and researchers, for
-preventing injection-style attacks. In this section, we defend the novelty and
-significance of our approach with respect to the state of the art in practice 
-and in research.
-
-Unlike frameworks provided by languages such as Haskell and
-Ruby, our type system provides a \emph{static} guarantee that input is always 
-properly sanitized before use. We achieve this by defining a typing relation
-which captures idiomatic sanitation algorithms. Type safety in our system relies upon 
-several closure and decidability results about operations on 
-regular languages which correspond to typical operations for sanitizing strings.
-
-Libraries and frameworks available in functional programming language communities often
-make claims about security and sometimes even mention sophisticated
-type systems as evidence of freedom from injection-style attacks.
-In reality, many of these approaches ultimately depend upon the correct implementation
-of library or framework input sanitation functions. In these cases, our type system
-provides a stronger (compile-time) guarantee and is additionally orthogonal because
-the correctness of the library or framework itself could be checked by implementing its
-sanitation functions in terms of our system.
-
-A number of research languages provide static guarantees that a program
-is free of input sanitation vulnerabilities. Most rely on some form of information
-flow TODO-nrf citations. Our extension to Ace differs from
-these systems in the ways following:
+The input sanitation problem is well-understood.
+There exist a large number of techniques and technologies, proposed by both practitioners and researchers, for preventing injection-style attacks.
+In this section, we explain how our approach to the input sanitation problem differs from each of these approaches.
+More important than these differences, however, is our more general assertion that language extensibility is a promising approach toward consideration of security goals in programming lanugage design.
+
+Unlike \emph{frameworks and libraries} provided by languages such as Haskell and Ruby, our type system provides a \emph{static} guarantee that input is always properly sanitized before use.
+Doing so requires reasoning about the operations on regular languages corresponding to standard operations on strings; we are unaware of any production system which contains this form of reasoning.
+Therefore, even where frameworks and libraries provide a viable interface or wrapper around input sanitation, our approach is complementary because it ensurees the correctness of the framework or library itself.
+Furthermore, our approach is more general than database abstraction layers because our mechanism is applicable to all forms of command injection (e.g.
+shell injection or remote file inclusion).
+
+A number of research languages provide static guarantees that a program is free of input sanitation vulnerabilities [Jif][Ur/Web]. Our work differes from these contributions in the ways following:
+
 \begin{itemize}
-  \item Our system is a light-weight solution to a single class of sanitation
-  vulnerabilities (e.g. we do not address Cross-Site Scripting). We present our
-  system not as a comprehensive solution to the web security problem, but rather as
-  evidence that composable, light-weight and simple analyses can address security
-  problems.
-  \item Our system is defined as a library in terms of an extensible type system,
-  as opposed to a stand-alone language. Instead of introducing new technologies
-  and methodologies for addressing security problems, we provide a light-weight
-  static analysis \emph{based upon existing approaches to the problem} which are
-  already well-understood by developers. We believe extensibility is a viable
-  and attractive alternative to highly specialized programming languages.
-  \item Ace is implemented in Python and shares its grammar. Since Python is a 
-  popular programming language among web developers, the barrier between our research
-  and adopted technologies is far lower than is e.g. Ur/Web's.
+  \item Our system is a light-weight solution to a single class of sanitation vulnerabilities (e.g. we do not address Cross-Site Scripting).
+  \item Our system is defined as a library in terms of an extensible type system, as opposed to a stand-alone language.
+Instead of introducing new technologies and methodologies for addressing security problems, we provide a light-weight static analysis which complements approaches developers already understand well.
+  \item Our implementation of the translation is implemented in Python and shares its grammar.
+Since Python is a popular programming language among web developers, the barrier between our research and adopted technologies is lower than for greenfield security-oriented languages.
 \end{itemize}
+
+We are also unaware of any extensible programming languages which emphasize applications to security concerns (TRUE?).
 
@@
 We differ from this and related work in at least two ways.
-First, our system  is defined within an extensible type system;
-second, and more importantly, we have demonstrated that regular expression types
-are applicable to the web security domain.
+First, our system  is defined within an extensible type system; second, and more importantly, we have demonstrated that regular expression types are applicable to the web security domain.
 
@@
   $\tconcat{P}{P}$ \alt
-  $\tsubst{P}{P}{P}$ \alt
+  $\preplace{P}{P}{P}$ \alt
   $\tcheck{P}{P}$ 

### 2014-08-05 sanitation-psp14/sanitation-psp14.tex
Added an abstract.

@@
 %Evidence suggests that programmers are reluctant to adopt new languages to
-ABSTRACT HERE
+Security-oriented programming languages do not enjoy wide adoption.
+Conversely, library and framework-based approaches to security are pervasive.
+Unfortunately, libraries and frameworks are only effective solutions if they are implemented and used properly.
+In this paper, we propose that extensible programming languages provide a promising mechanism for increasing the amount of trust that can be place in libraries and frameworks.
+To demonstrate this approach, we present a conservative extension to the simply-typed lambda calculus for checking the correctness of input sanitation algorithms.
+We also demonstrate how this language's security guarantees are preserved under translation into an underlying language containing a regular expression library.
+We believe this approach -- complementing existing techniques with light-weight, composable type-based analyses -- constitutes a promising compromise between usability, potential for industry adoption, and theoretically grounded safety guarantees.
 \end{abstract}
@@
 types ensuring a string belongs to a specified language. For instance, 
-$S : \rstr{r}$ reads ``$s$ is a string matching $r$".
+$S : \stringin{r}$ reads ``$s$ is a string matching $r$".
 the system includes an operation for replacing all instances of a pattern $r$ in a string $s_1$ with another string $s_2$. 

### 2014-08-05 sanitation-psp14/sanitation-psp14-short.tex
Adding short version.

--- 2014-08-05 sanitation-psp14/sanitation-psp14.tex
@@
 
-A number of research languages provide static guarantees that a program is free of input sanitation vulnerabilities [Jif][Ur/Web]. Our work differes from these contributions in the ways following:
-
-\begin{itemize}
-  \item Our system is a light-weight solution to a single class of sanitation vulnerabilities (e.g. we do not address Cross-Site Scripting).
-  \item Our system is defined as a library in terms of an extensible type system, as opposed to a stand-alone language.
+A number of research languages provide static guarantees that a program is free of input sanitation vulnerabilities [Jif][Ur/Web]. 
+Unlike this work, our solution to the input sanitation problem has a very low barrier to adoption; for instance, our implementation conservatively extends Python -- a popular language among web developers.
+We also believe our general approach is better-positioned for security, where continuously evolving threats might require frequent addition of new analyses; in these cases, the composability and generality of our approach is a substantial advantage.
+
+%\begin{itemize}
+%  \item Our system is a light-weight solution to a single class of sanitation vulnerabilities (e.g. we do not address Cross-Site Scripting).
+%  \item Our system is defined as a library in terms of an extensible type system, as opposed to a stand-alone language.
 Instead of introducing new technologies and methodologies for addressing security problems, we provide a light-weight static analysis which complements approaches developers already understand well.
-  \item Our implementation of the translation is implemented in Python and shares its grammar.
+%  \item Our implementation of the translation is implemented in Python and shares its grammar.
 Since Python is a popular programming language among web developers, the barrier between our research and adopted technologies is lower than for greenfield security-oriented languages.
-\end{itemize}
+%\end{itemize}
 
@@
 
-\subsection{Outline}
-
-An outline of this paper follows:
-
-\begin{itemize}
-  \item In \S 2, we define the type system which is embedded in Ace. We include a type safety proof for the string
-    segment of this language and prove the correctness of a translation to an underlying langguage $P$. In our theory,
-    $P$ is a simply typed lambda calculus equipped with a minimal regular expression library; in an implementation, $P$
-    stands in for Python or another underlying general-purpose programming language.
-  \item In \S 3, we discuss our implemention of this translation as a type system extension  within the Ace programming language.
-\end{itemize}
+%\subsection{Outline}
+%
+%An outline of this paper follows:
+%
+%\begin{itemize}
+%  \item In \S 2, we define the type system which is embedded in Ace. We include a type safety proof for the string
+%    segment of this language and prove the correctness of a translation to an underlying langguage $P$. In our theory,
+%    $P$ is a simply typed lambda calculus equipped with a minimal regular expression library; in an implementation, $P$
+%    stands in for Python or another underlying general-purpose programming language.
+%  \item In \S 3, we discuss our implemention of this translation as a type system extension  within the Ace programming language.
+%\end{itemize}
 
@@
 
-The system $\lambdas$ \ref{fig:lambdas} is the simply typed lambda calculus extended with \emph{regular expression types}, which are string
+The system $\lambdas$ is the simply typed lambda calculus extended with \emph{regular expression types}, which are string
 types ensuring a string belongs to a specified language. For instance, 
@@
 Input sanitation algorithms -- as implemented by developers or within popular libraries and frameworks -- are often implemented in terms of this replace operation.
-For instance, a developer might all potentially unsafe charapcters with excaped versions of the same character.
-Regular expression types are used both to specify input sanitation algorithms, and at use sites as specifications.
-Note that runtime error states ($S \err$) are introduced by coercion, not by replacement.
-
-The language $\lambdap$ \ref{fig:lambdap} is a simple functional language extended with a minimal regular expression library.
-Any general purpose programming language could stand in for $\lambdap$; for instance, SML has a regular expression library.
+
+The language $\lambdap$ is a simple functional language extended with a minimal regular expression library.
+Any general purpose programming language could stand in for $\lambdap$; for instance, SML or Python.
 In an implementation, our correctness results are modulo the underlying language's correct implementation of regular expression matching (see P-E-Replace).
 
-Finally, we define a translation from our type system $\lambdas$ into $\lambdap$.
-
+Finally, we define a translation from our type system $\lambdas$ into $\lambdap$ which preserves the safety guarnatee codified in the string types of $\lambdas$.
+Because our extension is conservative and its guarnatees are preserved under translation to an underlying language, it is highly composable with other analyses.
+
+Unfortunately, we are unable to present full proofs here due to space constraints.
 %Note that we never insert an internal check where the type of the string implies
@@
 
-Our type safety proofs for languages S and P and our translation correctness
-result all depend on some properties of regular languages. The crucial property
-is a relationship between string substitution -- which is available in any regular
-expression library -- and regular language substitution, which is a corresponding operation
-on languages instead of strings \ref{thm:substcorrespondence}.
-The decidability of language substitution is what enables static analysis of 
-sanitation algorithms implemented in terms of string replacement
-
+Our type safety proof for language S replies on a relationship between
+string substitution and language substitution given in lemma \ref{thm:substcorrespondence}.
+We also rely upon several other properties of regular languages.
 Throughout this section, we fix an alphabet $\Sigma$ over which strings $s$ and
@@
 The following are well-known properties of regular expressions which are necessary for our proofs:
-\begin{description}
-\item[(1):] If $s_1 \in \lang{r_1}$ and $s_2 \in \lang{r_2}$ then $s_1s_2 \in \lang{r_1r_2}$
-\item[(2):] For all strings $s$ and expressions $r$, either $s \in \lang{r}$ or $s \not \in \lang{r}$.
-\item[(3):] Regular languages are closed under complements and concatenation.
-\item[(4):] The regular expressions correspond bijectively to the regular languages.
-\end{description}
+If $s_1 \in \lang{r_1}$ and $s_2 \in \lang{r_2}$ then $s_1s_2 \in \lang{r_1r_2}$.
+For all strings $s$ and expressions $r$, either $s \in \lang{r}$ or $s \not \in \lang{r}$.
+Regular languages are closed under complements and concatenation.
+The regular expressions correspond bijectively to the regular languages.
 \end{lem}
@@
 \begin{lem}{Substitution Correspondence.} \label{thm:substcorrespondence}
-If $s_1 \in \lang{r_1}$ and $s_2 \in \lang{r_2}$ then $\lsubst{r}{s_1}{s_2} \in \lreplace{r}{r_1}{r_2}$.
+If $s_1 \in \lang{r_1}$ and $s_2 \in \lang{r_2}$ then $\lsubst{r}{s_1}{s_2} \in \lang{\lreplace{r}{r_1}{r_2}}$.
 \end{lem}
@@
 The proof proceeds by structural induction on $r$.
-
-\begin{description}
-\item[case $r=\alpha$.]
-If $s_1 = \alpha$ then $\alpha \in \lang{r_1}$ by assumption.
-Therefore, $\lsubst{r}{s_1}{s_2} = \lsubst{\alpha}{\alpha}{s_2} = s_2$ and
-$\lreplace{r}{r_1}{r_2} = \lreplace{\alpha}{r_1}{r_2}$. 
-Since $s_1 = \alpha$ and $\alpha \in \lang{r_1}$, $r_1 \iso \alpha | r_1'$ for some $r_1'$.
-Therefore, $\lreplace{\alpha}{r_1}{r_2} \iso \lreplace{\alpha}{\alpha|r_1'}{r_2}$ by Lemma X(1).
-Finally, $s_2 \in \lang{r_2}$ which implies $s_1 \in \lang{r_2|r'}$.
-If $s_1 \not = \alpha$ the $\lsubst{r}{s_1}{s_2} = s_1$ and $\lreplace{\alpha}{r_1}{r_2} = r_1$.
-
-\item[case $r = a|b$].
-Note that $\subst{a|b}{s_1}{s_2} = \lsubst{a}{\lsubst{b}{s_1}{s_2}}{s_2}$
-and $\lsubst{b}{s_1}{s_2} \in \lreplace{b}{r_1}{r_2}$ by induction.
-Therefore, $\lsubst{a|b}{s_1}{s_2}$ $\in \lreplace{a}{\lreplace{b}{r_1}{r_2}}{r_2}$ by induction and the definition of lreplace.
-Finally, applying definitions once more, $\lsubst{a|b}{s_1}{s_2} \in \lreplace{a|b}{r_1}{r_2}$.
-
-\item[case $r = ab$].
-By a similar argument to the disjunctive case.
-
-\item[case $r = a*$].
-By considering the once unwinding of $a*$, noting that $s_1$ and $s_2$ are finite.
-\end{description}
+The base case ($r=\alpha, \alpha \in \Sigma$) is trivial.
+The choice and sequential cases require lemma 1, and prove a more general property that $\lsubst{r}{s_1}{s_2} \in \lang{r'}$ for some $r'$ isomorphic to $\lreplace{r}{r_1}{r_2}$.
+The closure case follows from the induction hypothesis applied to any $n$-unwinding.
+
+%\begin{description}
+%\item[case $r=\alpha$.]
+%If $s_1 = \alpha$ then $\alpha \in \lang{r_1}$ by assumption.
+%Therefore, $\lsubst{r}{s_1}{s_2} = \lsubst{\alpha}{\alpha}{s_2} = s_2$ and
+%$\lreplace{r}{r_1}{r_2} = \lreplace{\alpha}{r_1}{r_2}$. 
+%Since $s_1 = \alpha$ and $\alpha \in \lang{r_1}$, $r_1 \iso \alpha | r_1'$ for some $r_1'$.
+%Therefore, $\lreplace{\alpha}{r_1}{r_2} \iso \lreplace{\alpha}{\alpha|r_1'}{r_2}$ by Lemma X(1).
+%Finally, $s_2 \in \lang{r_2}$ which implies $s_1 \in \lang{r_2|r'}$.
+%If $s_1 \not = \alpha$ the $\lsubst{r}{s_1}{s_2} = s_1$ and $\lreplace{\alpha}{r_1}{r_2} = r_1$.
+%
+%\item[case $r = a|b$].
+%Note that $\subst{a|b}{s_1}{s_2} = \lsubst{a}{\lsubst{b}{s_1}{s_2}}{s_2}$
+%and $\lsubst{b}{s_1}{s_2} \in \lreplace{b}{r_1}{r_2}$ by induction.
+%Therefore, $\lsubst{a|b}{s_1}{s_2} \in \lreplace{a}{\lreplace{b}{r_1}{r_2}}{r_2}$ by induction and the definition of lreplace.
+%Finally, applying definitions once more, $\lsubst{a|b}{s_1}{s_2} \in \lreplace{a|b}{r_1}{r_2}$.
+%
+%\item[case $r = ab$].
+%By a similar argument to the disjunctive case.
+%
+%\item[case $r = a*$].
+%By considering the once unwinding of $a*$, noting that $s_1$ and $s_2$ are finite.
+%\end{description}
 \end{proof}
@@
 \begin{proof}
-  The only non-trivial case is S-T-Replace, which follows from \ref{thm:total}. 
+  The only non-trivial case is S-T-Replace, which follows from lemma \ref{thm:total}. 
 \end{proof}
@@
 By induction on the derivation of $\sctx \vdash S : \psi$. The interesting case
-is S-T-Replace, which requires Lemma C.
-
-\begin{description}
-\item[S-T-Stringin-I:]
-If $S = \rsconcat{S_1}{S_2} : \stringin{r}$ then
-$S \sreduces S$ by S-E-RStr, and $\sctx \vdash S : \psi$ by assumption.
-
-\item[S-T-Concat:] 
-Suppose $S = \rsconcat{S_1}{S_2} : \stringin{r_1 r_2}$.
-By inversion, $\sctx \vdash S_1 : \stringin{r_1}$ and $\sctx \vdash S_1 : \stringin{r_2}$.
-It follows by induction that either $S_1 \err$, $S_2 \err$, 
-or $S_1 \sreduces \rstr{s_1}$ and $S_2 \sreduces \rstr{s_2}$ for some $s_1 \in \lang{r_1}$ and $s_2 \in \lang{r_2}$.
-In the latter case $S \reduces \rstr{s_1s_2}$ by S-E-Concat and $\sctx \vdash \rstr{s_1s_2} : \tstr{r_1r_2}$ by \ref{thm:regexprops}.
-In the former cases, $S \err$.
-
-\item[S-T-Replace:]
-Suppose $S = \rreplace{r}{S_1}{S_2}$ and $\sctx \vdash S : \stringin{r'}$.
-By inversion $\sctx \vdash S_1 : \stringin{r_1}$ and $\sctx \vdash S_2 : \stringin{r_2}$ such that $\lsubst{r}{r_1}{r_2} = r'$.
-By induction, $S_1 \err$, $S_2 \err$ 
-or $S_1 \sreduces \rstr{s_1}$ and $S_2 \sreduces \rstr{s_2}$ such that 
-In the latter case, we know $\lreplace{r}{s_1}{s_2} \in \lang{\lsubst{r}{r_1}{r_2}}$ by Lemma C;
-therefore by S-E-Replace, $S \sreduces \rstr{s}$ such that  $s \in \lang{\lsubst{r}{r_1}{r_2}} = \lang{r'}$.
-So by S-T-String-I, $\rstr{s} : \stringin{r'}$.
-In the former cases, $S \err$.
-
-\item[S-T-Coerce:]
-Suppose $S = \coerce{r}{S_1}$ and $S : \stringin{r}$.
-By inversion, $\sctx \vdash S_1 : \stringin{r'}$.
-By induction, $S_1 \err$ or $S_1 \sreduces \rstr{s}$.
-In the former case $S \err$ by propagation rules.
-In the latter case we have by property 2 of \ref{thm:regexprops} that $s \in \lang{r}$ or else
-$s \not \in \lang{r}$.
-If $s \in \lang{r}$ then $\rstr{s} : \stringin{r}$.
-If $s \not \in \lang{r}$ then $S \err$. 
-\end{description}
+is S-T-Replace, which requires lemma \ref{thm:substcorrespondence}. The 
+Coercion lemma additionally requires the second property of lemma \ref{thm:regexprops}.
+
+%\begin{description}
+%\item[S-T-Stringin-I:]
+%If $S = \rsconcat{S_1}{S_2} : \stringin{r}$ then
+%$S \sreduces S$ by S-E-RStr, and $\sctx \vdash S : \psi$ by assumption.
+%
+%\item[S-T-Concat:] 
+%Suppose $S = \rsconcat{S_1}{S_2} : \stringin{r_1 r_2}$.
+%By inversion, $\sctx \vdash S_1 : \stringin{r_1}$ and $\sctx \vdash S_1 : \stringin{r_2}$.
+%It follows by induction that either $S_1 \err$, $S_2 \err$, 
+%or $S_1 \sreduces \rstr{s_1}$ and $S_2 \sreduces \rstr{s_2}$ for some $s_1 \in \lang{r_1}$ and $s_2 \in \lang{r_2}$.
+%In the latter case $S \reduces \rstr{s_1s_2}$ by S-E-Concat and $\sctx \vdash \rstr{s_1s_2} : \tstr{r_1r_2}$ by \ref{thm:regexprops}.
+%In the former cases, $S \err$.
+%
+%\item[S-T-Replace:a]
+%Suppose $S = \rreplace{r}{S_1}{S_2}$ and $\sctx \vdash S : \stringin{r'}$.
+%By inversion $\sctx \vdash S_1 : \stringin{r_1}$ and $\sctx \vdash S_2 : \stringin{r_2}$ such that $\lsubst{r}{r_1}{r_2} = r'$.
+%By induction, $S_1 \err$, $S_2 \err$ 
+%or $S_1 \sreduces \rstr{s_1}$ and $S_2 \sreduces \rstr{s_2}$ such that 
+%In the latter case, we know $\lreplace{r}{s_1}{s_2} \in
+%\lang{\lsubst{r}{r_1}{r_2}}$ by lemma \ref{thm:substcorrespondence};
+%therefore by S-E-Replace, $S \sreduces \rstr{s}$ such that  
+%
+%$s \in \lang{\lsubst{r}{r_1}{r_2}} = \lang{r'}$.
+%So by S-T-String-I, $\rstr{s} : \stringin{r'}$.
+%In the former cases, $S \err$.
+%
+%\item[S-T-Coerce:]
+%Suppose $S = \coerce{r}{S_1}$ and $S : \stringin{r}$.
+%By inversion, $\sctx \vdash S_1 : \stringin{r'}$.
+%By induction, $S_1 \err$ or $S_1 \sreduces \rstr{s}$.
+%In the former case $S \err$ by propagation rules.
+%In the latter case we have by property 2 of \ref{thm:regexprops} that $s \in \lang{r}$ or else
+%$s \not \in \lang{r}$.
+%If $s \in \lang{r}$ then $\rstr{s} : \stringin{r}$.
+%If $s \not \in \lang{r}$ then $S \err$. 
+%\end{description}
 \end{proof}
@@
   Let $P$ be a term in the target language. If $\tctx \vdash P : \theta$ 
-  then $P \sreduces P'$ and $P' : \theta$.
+  then $P \sreduces P'$ and $\tctx \vdash P' : \theta$.
 \end{thm}
-\begin{proof}
-The proof proceeds by induction on the typing relation and is trivial give
-and inversion lemma for the typing relation. \textbf{We can write up this proof if we end up having enough space...}
-\end{proof}
 
@@
 If $S : \strin{r}$ then there exists a P such that $\trden{s} = P$ and either: 
-\begin{description}
-  \item[(a)] $P \treduces \tstr{s}$ and $S \sreduces \strin{s}$, and $s \in lang{r}$.
-  \item[(b)] $P \err$ and $S \err$.
-  \end{description}
+(a) $P \treduces \tstr{s}$ and $S \sreduces \strin{s}$, and $s \in lang{r}$; or
+(b) $P \err$ and $S \err$.
 \end{thm}
@@
 \begin{proof}
-The proof proceeds by induction on the typing relation for $S$.
+The proof proceeds by induction on the typing relation for $S$ and an appropriate
+choice of $P$; in each case, the choice is obvious. The subcases (a) proceed
+by inversion and appeals to the induction hypothesis. The subcases (b) proceed
+by the standard error propagation rules omitted for space.
 Throughout the proof, properties from the closure lemma for regular languages
-are necessary; for brevity, we elide these references.
-
-\begin{description}
-\item[S-T-String-I:] 
-Let $S = \rstr{s}$ and suppose $\sctx \vdash \rstr{s} : \stringin{r}$.
-Choose $T = \str{s}$ and note that $\trden{S} = P$ by Tr-String.
-By P-E-String, $P \treduces \str{s}$ and by S-E-String $S \treduces \rstr{s}$.
-Finally, by inversion of S-T-Stringin-I, $s \in \lang{r}$.
-
-\item[S-T-Concat:]
-Let $S = \rconcat{S_1}{S_2}$ and suppose $\sctx \vdash S : \stringin{r_1 r_2}$.
-By inversion, $\sctx \vdash S_1 : \stringin{r_1}$.
-It follows by induction that there exists a $P_1$ such that $\trden{S_1} = P_1$.
-By a similar argument for $S_2$ and $r_2$, there exists a $P_2$ such that $\trden{S_2} = P_2$.
-Choose $P = \concat{P_1}{P_2}$.
-
-We first prove property (a).
-Note that $S_1$ and $P_1$ are well typed (nrf ACTUALLY WE DON'T KNOW THAT $P_1$ IS WELL-TYPED!) and do not result in errors.
-Therefore, $S_1 \sreduces \rstr{s_1}$ and $P_1 \treduces \str{s_1}$ for some $s_1 \in \lang{r_1}$ by theorems 8 and 9 respectively. % TODO-nrf make sure this numbering is correct
-Similarly, $S_2 \sreduces \rstr{s_2}$ and $P_2 \treduces \str{s_2}$ for some $s_2 \in \lang{r_2}$.
-Therefore, $S \sreduces \rstr{s_1 s_2}$ by S-E-Concat and $\concat{P_1}{P_2} \treduces \str{s_1 s_2}$ by P-E-Concat.
-Finally, $s_1s_2 \in \lang{r_1}{r_2}$ by \ref{thm:regexprops}.
-
-Consider property (b).
-If $S_1 \err$ then $P_1 \err$ by induction, and it follows that $S \err$ and $P \err$ by respective error propagation rules.
-Similarly, if $S_2 \err$ then $P_2 \err$ and it follow that $S \err$ and $P \err$ by induction and propagation.
-
-\item[S-T-Replace:]
-Let $S = \rreplace{r}{S_1}{S_2}$ and suppose $\sctx \vdash S : \stringin{r'}$ for some $s$.
-By inversion of S-T-Replace, , $\sctx S_1 : \stringin{r_1}$ and $\sctx : \stringin{r_2}$ 
-such that $\lsubst{r}{r_1}{r_2} = r'$.
-By induction, there exists some $P_1, P_2$ such that $\trden{S_1} = P_1$, $\trden{S_2} = P_2$
-and either (a) or (b) holds.
-
-If (a) holds then $S_1 \sreduces \rstr{s_1}$ and $P_1 \treduces \str{s_1}$ for some $s_1 \in \lang{r_1}$,
-and similarly for $S_2, P_2$ and some $s_2 \in \lang{r_2}$.
-Therefore, by S-E-Replace, $S \sreduces \rstr{s}$ for some $s = \lreplace{r}{s_1}{s_2}$.
-Choose $P = \preplace{r}{s_1}{s_2}$.
-By a similar argument and P-E-Replace, $P \treduces \str{s}$ for some $s = \lreplace{r}{s_1}{s_2}$.
-What remains to be shown is $\lreplace{r}{s_1}{s_2} \in \lang{\lsubst{r}{r_1}{r_2}}$,
-which follows from Leamm D since $s_1 \in r_1$ and $s_2 \in r_2$.
-
-If (b) holds for $S_1$ and $P_1$, then $S \err$ and $P \err$ by propagation rules.
-Similarly, if (b) hols for $S_2$ and $P_2$ then $S \err$ and $P \err$ by propagation rules.
-
-\item[S-T-Coerce:]
-Let $S = \rcoerce{r}{S'}$ and suppose $\sctx \vdash \rcoerce{r}{S} : \stringin{r}$.
-By inverstion $\Psi \vdash S' : \stringin{r'}$ for an arbitrary $r'$.
-By induction there exists a $P'$ such that $\trden{S'} = P'$ and either (a) or (b)
-holds for $S'$ and $P'$.
-
-If (a) holds then $S' \sreduces \rstr{s'}$ and $P' \treduces \str{s'}$ for some $s' \in \lang{r'}$.
-Note that either $s' \in \lang{r}$ or $s' \not \in \lang{r}$ by property 2 of \ref{thm:regexprops}.
-Suppose $s' \in \lang{r}$. Then $\rcoerce{r}{S} \sreduces \rstr{S'}$ by S-E-Coerce.
-Choose $P = \check{\rx{r}}{P'}$ and note that $P \treduces \str{s'}$ by P-E-Coerce.
-Now suppose $s' \not \in \lang{r}$. Then $S \err$ and $P \err$ by P-E-Check-Err
-and S-E-Coerce-Err.
-
-Finally, if (b) holds then $S \err$ and $P \err$ by propagation.
-
-\end{description}
+are necessary.
+
+%\begin{description}
+%\item[S-T-String-I:] 
+%Let $S = \rstr{s}$ and suppose $\sctx \vdash \rstr{s} : \stringin{r}$.
+%Choose $T = \str{s}$ and note that $\trden{S} = P$ by Tr-String.
+%By P-E-String, $P \treduces \str{s}$ and by S-E-String $S \treduces \rstr{s}$.
+%Finally, by inversion of S-T-Stringin-I, $s \in \lang{r}$.
+%
+%\item[S-T-Concat:]
+%Let $S = \rconcat{S_1}{S_2}$ and suppose $\sctx \vdash S : \stringin{r_1 r_2}$.
+%By inversion, $\sctx \vdash S_1 : \stringin{r_1}$.
+%It follows by induction that there exists a $P_1$ such that $\trden{S_1} = P_1$.
+%By a similar argument for $S_2$ and $r_2$, there exists a $P_2$ such that $\trden{S_2} = P_2$.
+%Choose $P = \concat{P_1}{P_2}$.
+%
+%We first prove property (a).
+%Note that $S_1$ and $P_1$ do not result in errors and that $S_1$ is well typed.
+%Therefore, $S_1 \sreduces \rstr{s_1}$ and $P_1 \treduces \str{s_1}$ for some $s_1 \in \lang{r_1}$ by theorems 8 and 9 respectively. % TODO-nrf make sure this numbering is correct
+%Similarly, $S_2 \sreduces \rstr{s_2}$ and $P_2 \treduces \str{s_2}$ for some $s_2 \in \lang{r_2}$.
+%Therefore, $S \sreduces \rstr{s_1 s_2}$ by S-E-Concat and $\concat{P_1}{P_2} \treduces \str{s_1 s_2}$ by P-E-Concat.
+%Finally, $s_1s_2 \in \lang{r_1}{r_2}$ by \ref{thm:regexprops}.
+%
+%Consider property (b).
+%If $S_1 \err$ then $P_1 \err$ by induction, and it follows that $S \err$ and $P \err$ by respective error propagation rules.
+%Similarly, if $S_2 \err$ then $P_2 \err$ and it follow that $S \err$ and $P \err$ by induction and propagation.
+%
+%\item[S-T-Replace:]
+%Let $S = \rreplace{r}{S_1}{S_2}$ and suppose $\sctx \vdash S : \stringin{r'}$ for some $s$.
+%By inversion of S-T-Replace, , $\sctx S_1 : \stringin{r_1}$ and $\sctx : \stringin{r_2}$ 
+%such that $\lsubst{r}{r_1}{r_2} = r'$.
+%By induction, there exists some $P_1, P_2$ such that $\trden{S_1} = P_1$, $\trden{S_2} = P_2$
+%and either (a) or (b) holds.
+%
+%If (a) holds then $S_1 \sreduces \rstr{s_1}$ and $P_1 \treduces \str{s_1}$ for some $s_1 \in \lang{r_1}$,
+%and similarly for $S_2, P_2$ and some $s_2 \in \lang{r_2}$.
+%Therefore, by S-E-Replace, $S \sreduces \rstr{s}$ for some $s = \lreplace{r}{s_1}{s_2}$.
+%Choose $P = \preplace{r}{s_1}{s_2}$.
+%By a similar argument and P-E-Replace, $P \treduces \str{s}$ for some $s = \lreplace{r}{s_1}{s_2}$.
+%What remains to be shown is $\lreplace{r}{s_1}{s_2} \in \lang{\lsubst{r}{r_1}{r_2}}$,
+%which follows from Leamm D since $s_1 \in r_1$ and $s_2 \in r_2$.
+%
+%If (b) holds for $S_1$ and $P_1$, then $S \err$ and $P \err$ by propagation rules.
+%Similarly, if (b) hols for $S_2$ and $P_2$ then $S \err$ and $P \err$ by propagation rules.
+%
+%\item[S-T-Coerce:]
+%Let $S = \rcoerce{r}{S'}$ and suppose $\sctx \vdash \rcoerce{r}{S} : \stringin{r}$.
+%By inverstion $\Psi \vdash S' : \stringin{r'}$ for an arbitrary $r'$.
+%By induction there exists a $P'$ such that $\trden{S'} = P'$ and either (a) or (b)
+%holds for $S'$ and $P'$.
+%
+%If (a) holds then $S' \sreduces \rstr{s'}$ and $P' \treduces \str{s'}$ for some $s' \in \lang{r'}$.
+%Note that either $s' \in \lang{r}$ or $s' \not \in \lang{r}$ by property 2 of \ref{thm:regexprops}.
+%Suppose $s' \in \lang{r}$. Then $\rcoerce{r}{S} \sreduces \rstr{S'}$ by S-E-Coerce.
+%Choose $P = \check{\rx{r}}{P'}$ and note that $P \treduces \str{s'}$ by P-E-Coerce.
+%Now suppose $s' \not \in \lang{r}$. Then $S \err$ and $P \err$ by P-E-Check-Err
+%and S-E-Coerce-Err.
+%
+%Finally, if (b) holds then $S \err$ and $P \err$ by propagation.
+%
+%\end{description}
 \end{proof}
@@
 % the first empty line of section 2 and ending just before \end{document}.
+
+\section{Conclusion}
+
+Composable analyses which complement existing approaches constitute a promising approach toward the integration of security concerns into programming languages.
+In this paper, we presented a system with both of these properties and defined a security-prserving transformation.
+Unlike other approaches, our solution complements existing, familiar solutions while providing a strong guarantee that traditiona library and framework-based approaches are implemented and utilized correctly.
 

### 2014-08-05 sanitation-psp14/sanitation-psp14.tex
Added sec 3 back to the long version

@@
 
-%\newcommand{\F}[1]{\textsf{#1}~}
-%\newcommand{\FF}[1]{\textsf{#1}}
-%\newcommand{\Q}{\FF{Arg}}
-%\newcommand{\xlA}[1]{\lfloor #1 \rfloor_{\lambda}}
-%\newcommand{\xA}[1]{$\lfloor #1 \rfloor_{\text{Ace}}$}
-%\newcommand{\rtotau}[1]{\lfloor #1 \rfloor}
-%
-%\newcommand{\tremp}{{\tt tremp}}
-%\newcommand{\trdot}{{\tt trdot}}
-%\newcommand{\trchar}[1]{{\tt trchar}[#1]}
-%\newcommand{\trseq}[2]{{\tt trseq}(#1; #2)}
-%\newcommand{\tror}[2]{{\tt tror}(#1; #2)}
-%\newcommand{\tlstr}[1]{{\tt tstr}[#1]}
-%\begin{table*}[t]
-%\centering
-%\begin{tabular}{ l l l }
-%$r$ & $\xlA{r}$ & \xA{r}\\
-%\hline
-%$\epsilon$ & $\tremp$ & \verb|""|\\
-%$.$ & $\trdot$ & \verb|"."|\\
-%$a$ & $\trchar{a}$ & \verb|"|$a$\verb|"|\\
-%$r_1 \cdot r_2$ & $\trseq{\xlA{r_1}}{\xlA{r_2}}$ & \xA{r_1}\verb| + |\xA{r_2}\\
-%$r_1 + r_2$ & $\tror{\xlA{r_1}}{\xlA{r_2}}$ & \verb|"(" + |\xA{r_1}\verb! + "|" + !\xA{r_2}\verb| + ")"|\\
-%\\
-%$\psi$ & $\xlA{\psi}$ & \xA{\psi}\\
-%\hline
-%${\tt string\_in}[r]$ & $\fvar{stringin}[\xlA{r}]$ & \verb|string_in[|\xA{r}\verb|]|\\
-%\\
-%  $S$ & $\xlA{S}$ & \xA{S} \\
-%  \hline
-%  ${\tt str\_in}[s]$ & $\FF{intro}[\FF{str}[s]]()$ & \verb|"|$s$\verb|"|\\
-%  ${\tt concat}(S_1; S_2)$  & $\xlA{S_1}\cdot\FF{elim}[\tvar{concat}](\xlA{S_2})$ & \xA{S_1}\verb| + |\xA{S_2} \\
-%  ${\tt subst}[r](S_1; S_2)$  & $\xlA{S_1}\cdot\FF{elim}[\tvar{subst}~\xlA{r}](\xlA{S_2})$ & \xA{S_1}\verb|.subst(|\xA{r}\verb|, |\xA{S_2}\verb|)|\\
-%  ${\tt coerce}[r](S)$  & $\xlA{S}\cdot\FF{elim}[\tvar{coerce}~\xlA{r}]()$ & \xA{S}\verb|.coerce(|\xA{r}\verb|)|
-%\end{tabular}
-%\caption{Embeddings of the ${\tt string\_in}$ fragment into $\lamAce$ and Ace.}
-%\end{table*}
-%%
-%\newcommand{\atjsynX}[3]{\Gamma \vdash_\fvalCtx #1 \Rightarrow #2 \leadsto #3}
-%\newcommand{\atjanaX}[3]{\Gamma \vdash_\fvalCtx #1 \Leftarrow #2 \leadsto #3}
-%\newcommand{\atjerrX}[1]{\Gamma \vdash_\fvalCtx #1~ \mathtt{error}}
-%\begin{figure*}[t]
-%\small
-%$\fbox{\inferrule{}{\atjsynX{e}{\tau}{i}}}$~~~~
-%$\fbox{\inferrule{}{\atjanaX{e}{\tau}{i}}}$~~~~
-%%$\fbox{\inferrule{}{\atjerrX{e}}}$
-%\begin{mathpar}
-%\inferrule[att-flip]{
-%	\atjsynX{e}{\tau}{i}
+\newcommand{\F}[1]{\textsf{#1}~}
+\newcommand{\FF}[1]{\textsf{#1}}
+\newcommand{\Q}{\FF{Arg}}
+\newcommand{\xlA}[1]{\lfloor #1 \rfloor_{\lambda}}
+\newcommand{\xA}[1]{$\lfloor #1 \rfloor_{\text{Ace}}$}
+\newcommand{\rtotau}[1]{\lfloor #1 \rfloor}
+
+\newcommand{\tremp}{{\tt tremp}}
+\newcommand{\trdot}{{\tt trdot}}
+\newcommand{\trchar}[1]{{\tt trchar}[#1]}
+\newcommand{\trseq}[2]{{\tt trseq}(#1; #2)}
+\newcommand{\tror}[2]{{\tt tror}(#1; #2)}
+\newcommand{\tlstr}[1]{{\tt tstr}[#1]}
+\begin{table*}[t]
+\centering
+\begin{tabular}{ l l l }
+$r$ & $\xlA{r}$ & \xA{r}\\
+\hline
+$\epsilon$ & $\tremp$ & \verb|""|\\
+$.$ & $\trdot$ & \verb|"."|\\
+$a$ & $\trchar{a}$ & \verb|"|$a$\verb|"|\\
+$r_1 \cdot r_2$ & $\trseq{\xlA{r_1}}{\xlA{r_2}}$ & \xA{r_1}\verb| + |\xA{r_2}\\
+$r_1 + r_2$ & $\tror{\xlA{r_1}}{\xlA{r_2}}$ & \verb|"(" + |\xA{r_1}\verb! + "|" + !\xA{r_2}\verb| + ")"|\\
+\\
+$\psi$ & $\xlA{\psi}$ & \xA{\psi}\\
+\hline
+${\tt string\_in}[r]$ & $\fvar{stringin}[\xlA{r}]$ & \verb|string_in[|\xA{r}\verb|]|\\
+\\
+  $S$ & $\xlA{S}$ & \xA{S} \\
+  \hline
+  ${\tt str\_in}[s]$ & $\FF{intro}[\FF{str}[s]]()$ & \verb|"|$s$\verb|"|\\
+  ${\tt concat}(S_1; S_2)$  & $\xlA{S_1}\cdot\FF{elim}[\tvar{concat}](\xlA{S_2})$ & \xA{S_1}\verb| + |\xA{S_2} \\
+  ${\tt subst}[r](S_1; S_2)$  & $\xlA{S_1}\cdot\FF{elim}[\tvar{subst}~\xlA{r}](\xlA{S_2})$ & \xA{S_1}\verb|.subst(|\xA{r}\verb|, |\xA{S_2}\verb|)|\\
+  ${\tt coerce}[r](S)$  & $\xlA{S}\cdot\FF{elim}[\tvar{coerce}~\xlA{r}]()$ & \xA{S}\verb|.coerce(|\xA{r}\verb|)|
+\end{tabular}
+\caption{Embeddings of the ${\tt string\_in}$ fragment into $\lamAce$ and Ace.}
+\end{table*}
+%
+\newcommand{\atjsynX}[3]{\Gamma \vdash_\fvalCtx #1 \Rightarrow #2 \leadsto #3}
+\newcommand{\atjanaX}[3]{\Gamma \vdash_\fvalCtx #1 \Leftarrow #2 \leadsto #3}
+\newcommand{\atjerrX}[1]{\Gamma \vdash_\fvalCtx #1~ \mathtt{error}}
+\begin{figure*}[t]
+\small
+$\fbox{\inferrule{}{\atjsynX{e}{\tau}{i}}}$~~~~
+$\fbox{\inferrule{}{\atjanaX{e}{\tau}{i}}}$~~~~
+%$\fbox{\inferrule{}{\atjerrX{e}}}$
+\begin{mathpar}
+\inferrule[att-flip]{
+	\atjsynX{e}{\tau}{i}
+}{
+	\atjanaX{e}{\tau}{i}
+}
+
+%\inferrule[att-var]{
+%	x \Rightarrow \tau \in \Gamma
 %}{
-%	\atjanaX{e}{\tau}{i}
+%	\atjsynX{x}{\tau}{x}
 %}
 %
-%%\inferrule[att-var]{
-%%	x \Rightarrow \tau \in \Gamma
-%%}{
-%%	\atjsynX{x}{\tau}{x}
-%%}
-%%
-%%\inferrule[att-asc]{
-%%    \tau \Downarrow_\fvalCtx \tau'\\
-%%	\atjanaX{e}{\tau'}{i}
-%%}{
-%%	\atjsynX{e : \tau}{\tau'}{i}
-%%}
-%%
-%%\inferrule[att-let-syn]{
-%%	\atjsynX{e_1}{\tau_1}{i_1}\\
-%%	\Gamma, x \Rightarrow \tau_1 \vdash_\fvalCtx e_2 \Rightarrow \tau_2 \leadsto i_2\\
-%%	\trepof{\tau_1} \Downarrow_\fvalCtx \titype{\sigma_1}
-%%}{
-%%	\atjsynX{\F{let}x = e_1~\F{in}e_2}{\tau_2}{(\ilam{x}{\sigma_1}{i_2})~i_1}
-%%}
-%%
-%%\inferrule[att-lam-ana]{
-%%	\Gamma, x \Rightarrow \tau_1 \vdash_\fvalCtx e \Leftarrow \tau_2 \leadsto i\\
-%%	\trepof{\tau_1} \Downarrow_\fvalCtx \titype{\sigma_1}
-%%}{
-%%	\atjanaX{\lambda x.e}{\ttype{arrow}{(\tau_1, \tau_2)}}{\ilam{x}{\sigma_1}{i}}
-%%}
-%%
-%%\inferrule[att-lam]{
-%%	\tau_1 \Downarrow_\fvalCtx \tau_1'\\
-%%		\trepof{\tau_1'} \Downarrow_\fvalCtx \titype{\sigma}\\\\
-%%	\Gamma, x \Rightarrow \tau_1' \vdash_\fvalCtx e \Rightarrow \tau_2 \leadsto i
-%%%	i \hookrightarrow_\fvalCtx i'\\
-%%%		\sigma \hookrightarrow_\fvalCtx \sigma'
-%%%	\ddbar{\fvar{Arrow}}{\fvalCtx}{\trepof{\tau_1'}}{\sbar_1}\\
-%%	%\delfromtau{$\Xi_0$}{\fvalCtx}{\tau_1'}{\sabs}\\\\
-%%}{
-%%	\atjsynX{\elam{x}{\tau_1}{e}}{\ttype{arrow}{(\tau_1', \tau_2)}}{\ilam{x}{\sigma'}{i}}
-%%}
-%%
-%\inferrule[att-intro-ana]{
-%	\vdash_\fvalCtx \FF{iana}(\fvar{tycon})=\taudef\\
-%	\taudef~\taut{opidx}~\taut{tyidx}~((\Gamma; e_1)? :: \ldots :: (\Gamma; e_n)? :: []) \Downarrow_\fvalCtx \titerm{i}\\\\
-%%	\trepof{\ttype{tycon}{\tauidx'}} \Downarrow_\fvalCtx \titype{\sigma}\\
-%		\trepof{\ttype{tycon}{\taut{tyidx}}} \Downarrow_\fvalCtx \titype{\sigma}\\
+%\inferrule[att-asc]{
+%    \tau \Downarrow_\fvalCtx \tau'\\
+%	\atjanaX{e}{\tau'}{i}
+%}{
+%	\atjsynX{e : \tau}{\tau'}{i}
+%}
+%
+%\inferrule[att-let-syn]{
+%	\atjsynX{e_1}{\tau_1}{i_1}\\
+%	\Gamma, x \Rightarrow \tau_1 \vdash_\fvalCtx e_2 \Rightarrow \tau_2 \leadsto i_2\\
+%	\trepof{\tau_1} \Downarrow_\fvalCtx \titype{\sigma_1}
+%}{
+%	\atjsynX{\F{let}x = e_1~\F{in}e_2}{\tau_2}{(\ilam{x}{\sigma_1}{i_2})~i_1}
+%}
+%
+%\inferrule[att-lam-ana]{
+%	\Gamma, x \Rightarrow \tau_1 \vdash_\fvalCtx e \Leftarrow \tau_2 \leadsto i\\
+%	\trepof{\tau_1} \Downarrow_\fvalCtx \titype{\sigma_1}
+%}{
+%	\atjanaX{\lambda x.e}{\ttype{arrow}{(\tau_1, \tau_2)}}{\ilam{x}{\sigma_1}{i}}
+%}
+%
+%\inferrule[att-lam]{
+%	\tau_1 \Downarrow_\fvalCtx \tau_1'\\
+%		\trepof{\tau_1'} \Downarrow_\fvalCtx \titype{\sigma}\\\\
+%	\Gamma, x \Rightarrow \tau_1' \vdash_\fvalCtx e \Rightarrow \tau_2 \leadsto i
+%%	i \hookrightarrow_\fvalCtx i'\\
+%%		\sigma \hookrightarrow_\fvalCtx \sigma'
+%%	\ddbar{\fvar{Arrow}}{\fvalCtx}{\trepof{\tau_1'}}{\sbar_1}\\
+%	%\delfromtau{$\Xi_0$}{\fvalCtx}{\tau_1'}{\sabs}\\\\
+%}{
+%	\atjsynX{\elam{x}{\tau_1}{e}}{\ttype{arrow}{(\tau_1', \tau_2)}}{\ilam{x}{\sigma'}{i}}
+%}
+%
+\inferrule[att-intro-ana]{
+	\vdash_\fvalCtx \FF{iana}(\fvar{tycon})=\taudef\\
+	\taudef~\taut{opidx}~\taut{tyidx}~((\Gamma; e_1)? :: \ldots :: (\Gamma; e_n)? :: []) \Downarrow_\fvalCtx \titerm{i}\\\\
+%	\trepof{\ttype{tycon}{\tauidx'}} \Downarrow_\fvalCtx \titype{\sigma}\\
+		\trepof{\ttype{tycon}{\taut{tyidx}}} \Downarrow_\fvalCtx \titype{\sigma}\\
+		\Gamma \vdash_\fvalCtx i : \sigma
+}{
+	\atjanaX{\FF{intro}[\taut{opidx}](e_1; \ldots; e_n)}{\ttype{tycon}{\taut{tyidx}}}{i}
+}
+
+%\inferrule[att-i-asc-ty]{
+%	\atjanaX{I}{\tau}{i}
+%}{
+%	\atjsynX{I : \tau}{\tau}{i}
+%}
+%
+%\inferrule[att-i-asc-tycon]{
+%	\vdash_\fvalCtx \FF{isyn}(\fvar{tycon})=\taudef\\\\
+%	\taudef~\tauidx~((\Gamma; e_1)? {::}{\ldots}{::}(\Gamma; e_n)? {::} []) \Downarrow_\fvalCtx \tden{\titerm{i}}{\ttype{tycon}{\tauidx'}}\\
+%			\trepof{\tau_1'} \Downarrow_\fvalCtx \titype{\sigma}\\
 %	\Gamma \vdash_\fvalCtx i : \sigma
 %}{
-%	\atjanaX{\FF{intro}[\taut{opidx}](e_1; \ldots; e_n)}{\ttype{tycon}{\taut{tyidx}}}{i}
+%	\atjsynX{\FF{intro}[\tauidx](e_1; \ldots; e_n)] :: \fvar{tycon}}{\ttype{tycon}{\tauidx'}}{i'}
 %}
 %
-%%\inferrule[att-i-asc-ty]{
-%%	\atjanaX{I}{\tau}{i}
-%%}{
-%%	\atjsynX{I : \tau}{\tau}{i}
-%%}
-%%
-%%\inferrule[att-i-asc-tycon]{
-%%	\vdash_\fvalCtx \FF{isyn}(\fvar{tycon})=\taudef\\\\
-%%	\taudef~\tauidx~((\Gamma; e_1)? {::}{\ldots}{::}(\Gamma; e_n)? {::} []) \Downarrow_\fvalCtx \tden{\titerm{i}}{\ttype{tycon}{\tauidx'}}\\
-%%			\trepof{\tau_1'} \Downarrow_\fvalCtx \titype{\sigma}\\
-%%	\Gamma \vdash_\fvalCtx i : \sigma
-%%}{
-%%	\atjsynX{\FF{intro}[\tauidx](e_1; \ldots; e_n)] :: \fvar{tycon}}{\ttype{tycon}{\tauidx'}}{i'}
-%%}
-%%
-%\inferrule[att-elim-syn]{
-%	\atjsynX{e}{\ttype{tycon}{\taut{tyidx}}}{i}\\
-%	\vdash_\fvalCtx \FF{esyn}(\fvar{tycon})=\taudef\\\\
-%	\taudef~\taut{opidx}~\taut{tyidx}~\titerm{i}~((\Gamma; e)? :: (\Gamma; e_1)? :: \ldots :: (\Gamma; e_n)? :: []) \Downarrow_\fvalCtx (\tau, \titerm{i'})\\\\
-%			\trepof{\tau} \Downarrow_\fvalCtx \titype{\sigma}\\
-%	\Gamma \vdash_\fvalCtx i' : \sigma
+\inferrule[att-elim-syn]{
+	\atjsynX{e}{\ttype{tycon}{\taut{tyidx}}}{i}\\
+	\vdash_\fvalCtx \FF{esyn}(\fvar{tycon})=\taudef\\\\
+	\taudef~\taut{opidx}~\taut{tyidx}~\titerm{i}~((\Gamma; e)? :: (\Gamma; e_1)? :: \ldots :: (\Gamma; e_n)? :: []) \Downarrow_\fvalCtx (\tau, \titerm{i'})\\\\
+			\trepof{\tau} \Downarrow_\fvalCtx \titype{\sigma}\\
+	\Gamma \vdash_\fvalCtx i' : \sigma
+}{
+	\atjsynX{e\cdot\FF{elim}[\taut{opidx}](e_1; \ldots; e_n)}{\tau}{i'}
+}
+\end{mathpar}
+%\vspace{-10px}
+\caption{The bidirectional active typechecking and translation judgements.}
+\label{atj}
+\end{figure*}
+\newcommand{\tlevalX}[2]{#1 \Downarrow_\fvalCtx #2}
+\begin{figure*}[t]
+\small
+$\fbox{\inferrule{}{\tau \Downarrow_\fvalCtx \tau'}}$~~~~
+\begin{mathpar}
+\inferrule[trstr-eval]{ }{\tlevalX{\tlstr{s}}{\tlstr{s}}}
+
+\inferrule[tremp-eval]{ }{\tremp \Downarrow_\fvalCtx \tremp}
+
+\cdots
+
+\inferrule[tror-eval]{
+	\tlevalX{\tau_1}{\tau_1'}\\
+	\tlevalX{\tau_2}{\tau_2'}
+}{
+	\tlevalX{\tror{\tau_1}{\tau_2}}{\tror{\tau_1'}{\tau_2'}}
+}
+
+trmatch
+
+trlsubst
+
+trsublang
+
+%\inferrule[repof]{
+%	\tau \Downarrow_\fvalCtx \ttype{tycon}{\tauidx}\\
+%	\vdash_\fvalCtx \FF{rep}(\fvar{tycon}) = \taurep\\
+%	\taurep~\tauidx \Downarrow_\fvalCtx \titype{\sigma}
 %}{
-%	\atjsynX{e\cdot\FF{elim}[\taut{opidx}](e_1; \ldots; e_n)}{\tau}{i'}
+%	\FF{repof}(\tau) \Downarrow_\fvalCtx \titype{\sigma}
 %}
-%\end{mathpar}
-%%\vspace{-10px}
-%\caption{The bidirectional active typechecking and translation judgements.}
-%\label{atj}
-%\end{figure*}
-%\newcommand{\tlevalX}[2]{#1 \Downarrow_\fvalCtx #2}
-%\begin{figure*}[t]
-%\small
-%$\fbox{\inferrule{}{\tau \Downarrow_\fvalCtx \tau'}}$~~~~
-%\begin{mathpar}
-%\inferrule[trstr-eval]{ }{\tlevalX{\tlstr{s}}{\tlstr{s}}}
-%
-%\inferrule[tremp-eval]{ }{\tremp \Downarrow_\fvalCtx \tremp}
-%
-%\cdots
-%
-%\inferrule[tror-eval]{
-%	\tlevalX{\tau_1}{\tau_1'}\\
-%	\tlevalX{\tau_2}{\tau_2'}
-%}{
-%	\tlevalX{\tror{\tau_1}{\tau_2}}{\tror{\tau_1'}{\tau_2'}}
-%}
-%
-%trmatch
-%
-%trlsubst
-%
-%trsublang
-%
-%
-%%\inferrule[repof]{
-%%	\tau \Downarrow_\fvalCtx \ttype{tycon}{\tauidx}\\
-%%	\vdash_\fvalCtx \FF{rep}(\fvar{tycon}) = \taurep\\
-%%	\taurep~\tauidx \Downarrow_\fvalCtx \titype{\sigma}
-%%}{
-%%	\FF{repof}(\tau) \Downarrow_\fvalCtx \titype{\sigma}
-%%}
-%%
-%\inferrule[syn]{
-%    \tau \Downarrow_\fvalCtx (\Gamma; e)?\\
-%	\atjsynX{e}{\tau}{\iota}\\\\
-%    [\tau/\tvar{t}_{ty}, \titerm{\iota}/\tvar{t}_{trans}]\tau_2 \Downarrow_\fvalCtx \tau_2'
-%}{
-%	\FF{syn}(\tau_1; \tvar{t}_{ty}, \tvar{t}_{trans}.\tau_2) \Downarrow_\fvalCtx \tau_2'
-%}
-%
-%\inferrule[ana]{
-%	\tau_1 \Downarrow_\fvalCtx (\Gamma; e)?\\
-%	\tau_2 \Downarrow_\fvalCtx \tau_2'\\
-%	\atjanaX{e}{\tau_2'}{\iota}\\\\
-%	[\titerm{\iota}/\tvar{t}_{trans}]\tau_3 \Downarrow_\fvalCtx \tau_3'
-%}{
-%	\FF{ana}(\tau_1; \tau_2; \tvar{t}_{trans}.\tau_3) \Downarrow_\fvalCtx \tau_3'
-%}
-%
-%TODO: add rx, str stuff
-%\end{mathpar}
-%\caption{\small Normalization semantics for the type-level language.
+%
+\inferrule[syn]{
+    \tau \Downarrow_\fvalCtx (\Gamma; e)?\\
+	\atjsynX{e}{\tau}{\iota}\\\\
+    [\tau/\tvar{t}_{ty}, \titerm{\iota}/\tvar{t}_{trans}]\tau_2 \Downarrow_\fvalCtx \tau_2'
+}{
+	\FF{syn}(\tau_1; \tvar{t}_{ty}, \tvar{t}_{trans}.\tau_2) \Downarrow_\fvalCtx \tau_2'
+}
+
+\inferrule[ana]{
+	\tau_1 \Downarrow_\fvalCtx (\Gamma; e)?\\
+	\tau_2 \Downarrow_\fvalCtx \tau_2'\\
+	\atjanaX{e}{\tau_2'}{\iota}\\\\
+	[\titerm{\iota}/\tvar{t}_{trans}]\tau_3 \Downarrow_\fvalCtx \tau_3'
+}{
+	\FF{ana}(\tau_1; \tau_2; \tvar{t}_{trans}.\tau_3) \Downarrow_\fvalCtx \tau_3'
+}
+
+TODO: add rx, str stuff
+\end{mathpar}
+\caption{\small Normalization semantics for the type-level language.
 Missing rules (including error propagation rules and normalization of quoted internal terms and types) are unsurprising and will be given later.}
-%\label{tleval}
-%\end{figure*}
-%\begin{figure}[t]
-%\small\begin{flalign}
-%& \F{tycon}\fvar{stringin}~\F{of}\FF{R}~\{\\
-%& \quad \F{iana}\{\tlam{opidx}{\FF{String}}{
-%	\tlam{tyidx}{\FF{R}}{
-%	\tlam{a}{\klist{\Q}}{\\
-%& \quad\quad \tvar{arity0}~\tvar{a}~(\tvar{check}~\tvar{opidx}~\tvar{tyidx}~\titerm{\iup{\tvar{opidx}}})
-%	}}
-%}\}\\
-%& \quad \F{esyn}\{\tlam{opidx}{\kunit+(\FF{R}+\FF{R})}{
-%	\tlam{a}{\klist{\Q}}{\\
-%& \quad\quad \tsumcase{\tvar{opidx}}{\_}{\\
-%& \quad\quad\quad \tvar{arity2}~\tvar{a}~\tlam{a1}{\Q}{\tlam{a2}{\Q}{\\
-%& \quad\quad\quad\quad \tvar{rsyn}~\tvar{a1}~\tlam{r1}{\FF{R}}{\tlam{i1}{\kITerm}{~}}\\
-%& \quad\quad\quad\quad \tvar{rsyn}~\tvar{a2}~\tlam{r2}{\FF{R}}{\tlam{i2}{\kITerm}{~}}\\
-%& \quad\quad\quad\quad\quad (\ttype{stringin}{\FF{rseq}(\tvar{r1}; \tvar{r2})}, \\
-%& \quad\quad\quad\quad\quad\titerm{{\tt iconcat}(\iup{\tvar{i1}}; \iup{\tvar{i2}})})
-%}}\\
-%& \quad\quad}{opidx'}{d}}
-%}\}
-%\};\\
-%& \F{def}\tvar{concat} = \tinl{\kunit, \FF{R}+\FF{R}}{\tunit};\\
-%& \F{def}\tvar{subst} = \tlam{r}{\FF{R}}{\tinr{\kunit, \FF{R}+\FF{R}}{\tinl{\FF{R},\FF{R}}{\tvar{r}}}};\\
-%& \F{def}\tvar{coerce} = \tlam{r}{\FF{R}}{\tinr{\kunit, \FF{R}+\FF{R}}{\tinr{\FF{R},\FF{R}}{\tvar{r}}}}
-%\end{flalign}
-%\caption{Definition of $\phi_S$, which enables the embedding of fragment $S$ into $\lamAce$.}
+\label{tleval}
+\end{figure*}
+\begin{figure}[t]
+\small\begin{flalign}
+& \F{tycon}\fvar{stringin}~\F{of}\FF{R}~\{\\
+& \quad \F{iana}\{\tlam{opidx}{\FF{String}}{
+	\tlam{tyidx}{\FF{R}}{
+	\tlam{a}{\klist{\Q}}{\\
+& \quad\quad \tvar{arity0}~\tvar{a}~(\tvar{check}~\tvar{opidx}~\tvar{tyidx}~\titerm{\iup{\tvar{opidx}}})
+	}}
+}\}\\
+& \quad \F{esyn}\{\tlam{opidx}{\kunit+(\FF{R}+\FF{R})}{
+	\tlam{a}{\klist{\Q}}{\\
+& \quad\quad \tsumcase{\tvar{opidx}}{\_}{\\
+& \quad\quad\quad \tvar{arity2}~\tvar{a}~\tlam{a1}{\Q}{\tlam{a2}{\Q}{\\
+& \quad\quad\quad\quad \tvar{rsyn}~\tvar{a1}~\tlam{r1}{\FF{R}}{\tlam{i1}{\kITerm}{~}}\\
+& \quad\quad\quad\quad \tvar{rsyn}~\tvar{a2}~\tlam{r2}{\FF{R}}{\tlam{i2}{\kITerm}{~}}\\
+& \quad\quad\quad\quad\quad (\ttype{stringin}{\FF{rseq}(\tvar{r1}; \tvar{r2})}, \\
+& \quad\quad\quad\quad\quad\titerm{{\tt iconcat}(\iup{\tvar{i1}}; \iup{\tvar{i2}})})
+}}\\
+& \quad\quad}{opidx'}{d}}
+}\}
+\};\\
+& \F{def}\tvar{concat} = \tinl{\kunit, \FF{R}+\FF{R}}{\tunit};\\
+& \F{def}\tvar{subst} = \tlam{r}{\FF{R}}{\tinr{\kunit, \FF{R}+\FF{R}}{\tinl{\FF{R},\FF{R}}{\tvar{r}}}};\\
+& \F{def}\tvar{coerce} = \tlam{r}{\FF{R}}{\tinr{\kunit, \FF{R}+\FF{R}}{\tinr{\FF{R},\FF{R}}{\tvar{r}}}}
+\end{flalign}
+\caption{Definition of $\phi_S$, which enables the embedding of fragment $S$ into $\lamAce$.}
+\end{figure}
+%\begin{figure}
+%\[
+%\begin{array}{lcl}
+%% & & Definition & Kind\\
+%\tvar{concat} & := & \tinl{\kunit, \FF{R}+\FF{R}}{\tunit}\\%& \kunit + (R + R)\\
+%\tvar{replace} & := & \tlam{r}{\FF{R}}{\tinr{\kunit, \FF{R}+\FF{R}}{\tinl{\FF{R},\FF{R}}{\tvar{r}}}}\\% & \karrow{a}{b}\\
+%\tvar{coerce} & := & \tlam{r}{\FF{R}}{\tinr{\kunit, \FF{R}+\FF{R}}{\tinr{\FF{R},\FF{R}}{\tvar{r}}}}\\
+%\end{array}
+%\]
+%\caption{Definitions}
 %\end{figure}
-%%\begin{figure}
-%%\[
-%%\begin{array}{lcl}
-%%% & & Definition & Kind\\
-%%\tvar{concat} & := & \tinl{\kunit, \FF{R}+\FF{R}}{\tunit}\\%& \kunit + (R + R)\\
-%%\tvar{replace} & := & \tlam{r}{\FF{R}}{\tinr{\kunit, \FF{R}+\FF{R}}{\tinl{\FF{R},\FF{R}}{\tvar{r}}}}\\% & \karrow{a}{b}\\
-%%\tvar{coerce} & := & \tlam{r}{\FF{R}}{\tinr{\kunit, \FF{R}+\FF{R}}{\tinr{\FF{R},\FF{R}}{\tvar{r}}}}\\
-%%\end{array}
-%%\]
-%%\caption{Definitions}
-%%\end{figure}
-%
-%\subsection{Background: Ace}
-%TODO: Make a TR out of the OOPSLA submission.
-%\subsection{Explicit Conversions}
-%\subsection{Adding Subtyping to Ace}
-%\subsection{Theory}
-%
-%\section{Related Work}
-%
-%\section{Discussion}
-%
-%
-%\bibliographystyle{abbrv}
-%
-%% The bibliography should be embedded for final submission.
-%
-%\bibliography{../research}
-%%\softraggedright
-%%P. Q. Smith, and X. Y. Jones. ...reference text...
-%
-%
+
+\subsection{Background: Ace}
+TODO: Make a TR out of the OOPSLA submission.
+\subsection{Explicit Conversions}
+\subsection{Adding Subtyping to Ace}
+\subsection{Theory}
+
+\section{Related Work}
+
+\section{Discussion}
+
+\bibliographystyle{abbrv}
+
+% The bibliography should be embedded for final submission.
+
+\bibliography{../research}
+%\softraggedright
+%P. Q. Smith, and X. Y. Jones. ...reference text...
 

### 2014-08-05 sanitation-psp14/sanitation-psp14-short.tex
elements can't be iso

@@
 The base case ($r=\alpha, \alpha \in \Sigma$) is trivial.
-The choice and sequential cases require lemma 1, and prove a more general property that $\lsubst{r}{s_1}{s_2} \in \lang{r'}$ for some $r'$ isomorphic to $\lreplace{r}{r_1}{r_2}$.
+The choice and sequential cases require lemma 1, and prove a more general property that $\lsubst{r}{s_1}{s_2} \in \lang{r'}$ for some $r'$ equivalent to $\lreplace{r}{r_1}{r_2}$.
 The closure case follows from the induction hypothesis applied to any $n$-unwinding.

### 2014-08-05 sanitation-psp14/sanitation-psp14-short.tex
Fixed typos.

@@
 When these methods are unavailable or insufficient, developers implement custom sanitation techniques.
-Im both cases, sanitation algorithms are implemented using the language's regular expression capabilities and usually \emph{replace} potentially unsafe strings with equivalent escaped strings.
+In both cases, sanitation algorithms are implemented using the language's regular expression capabilities and usually \emph{replace} potentially unsafe strings with equivalent escaped strings.
 
@@
 Libraries, frameworks or functions which construct and execute commands containing input can specify a safe subset $\stringin{r_{\sf spec}}$ of strings, and input sanitation algorithms can construct such a string using ${\sf rreplace}$ or, optionally, by coercion (in which case a runtime check is inserted).
-We also show how this simple can be translated into a host language containing a regular expression library such that the safety guarantee of the extended language is preserved.
+We also show how this system is translated into a host language containing a regular expression library such that the safety guarantee of the extended language is preserved.
 
@@
 
-Unfortunately, we are unable to present full proofs here due to space constraints.
+Unfortunately, we are unable to present full proofs in this paper due to space constraints; however, proofs are given in full in the accompanying technical report?.
 %Note that we never insert an internal check where the type of the string implies
@@
 choice of $P$; in each case, the choice is obvious. The subcases (a) proceed
-by inversion and appeals to the induction hypothesis. The subcases (b) proceed
+by inversion and appeals to our type safety theorems as well as the induction hypothesis. The subcases (b) proceed
 by the standard error propagation rules omitted for space.
@@
 Composable analyses which complement existing approaches constitute a promising approach toward the integration of security concerns into programming languages.
-In this paper, we presented a system with both of these properties and defined a security-prserving transformation.
-Unlike other approaches, our solution complements existing, familiar solutions while providing a strong guarantee that traditiona library and framework-based approaches are implemented and utilized correctly.
+In this paper, we presented a system with both of these properties and defined a security-preserving transformation.
+Unlike other approaches, our solution complements existing, familiar solutions while providing a strong guarantee that traditional library and framework-based approaches are implemented and utilized correctly.
 

### 2014-08-05 sanitation-psp14/sanitation-psp14.tex
Fixed typos.

@@
 Additionally, we include an operation $\rreplace{r}{s_1}{s_2}$ which corresponds to the replace mechanism available in most regular expression libraries; that is, any substring of $s_1$ matching $r$ is replaced with $s_2$.
-The type of this expression is the computed, and is likely ``smaller" or more constrained than the type of $s_1$.
-Libraries, frameworks or functions which construct and execute commands containing input can specify a safe subset $\stringin{r_{\sf spec}}$ of strings, and input sanitation algorithms can construct such a string using ${\sf rreplace}$ or, optionally, by coercion (in which case a runtime check is inserted).
-We also show how this simple can be translated into a host language containing a regular expression library such that the safety guarantee of the extended language is preserved.
+The type of this expression is computed, and is likely ``smaller" or more constrained than the type of $s_1$.
+Libraries, frameworks or functions which construct and execute commands containing input can specify a safe subset $\stringin{r_{\sf spec}}$ of strings, and input sanitation algorithms can construct a string with this type using ${\sf rreplace}$ or, optionally, by coercion (in which case a runtime check is inserted).
+We also show how this simple extension can be translated into a host language containing a regular expression library such that the safety guarantee of the extended language is preserved.
 

### 2014-08-05 sanitation-psp14/sanitation-psp14-short.tex
Improved description of proof a bit.

@@
 If $S : \strin{r}$ then there exists a P such that $\trden{s} = P$ and either: 
-(a) $P \treduces \tstr{s}$ and $S \sreduces \strin{s}$, and $s \in lang{r}$; or
+(a) $P \treduces \tstr{s}$ and $S \sreduces \strin{s}$, and $s \in \lang{r}$; or
 (b) $P \err$ and $S \err$.

### 2014-08-06 sanitation-psp14/sanitation-psp14-short.tex
Updated PSP14 abstract

--- 2014-08-05 sanitation-psp14/sanitation-psp14-short.tex
@@
-\documentclass{acm_proc_article-sp}
+\documentclass[9pt]{sig-alternate}
 \newcommand{\lamAce}{\lambda_{\text{Ace}}}
@@
 
-\title{A Type System for String Sanitation\\Implemented Inside a Python}
-
+\title{Statically Typed String Sanitation Inside a Python}
+\numberofauthors{3}
+\author{
+  \alignauthor
+  Nathan Fulton
+  \alignauthor
+  Cyrus Omar
+  \alignauthor
+  Jonathan Aldrich
+  \and
+  \affaddr{Carnegie Mellon University}\\
+  \affaddr{Pittsburgh, PA}\\
+  \email{\{nathanfu, comar, aldrich\}@cs.cmu.edu}
+}
 %\authorinfo{~}{~}{~}
-%dep_docclass\authorinfo{Nathan Fulton\and Cyrus Omar\and Jonathan Aldrich}
-%dep_docclass           {School of Computer Science\\
-%dep_docclass            Carnegie Mellon University}
-%dep_docclass           {\{nathanfu, comar, aldrich\}@cs.cmu.edu}
+%\authorinfo{Nathan Fulton\and Cyrus Omar\and Jonathan Aldrich}
+ %          {School of Computer Science\\
+  %          Carnegie Mellon University
+   %        \{nathanfu, comar, aldrich\}@cs.cmu.edu}
 
@@
 %Evidence suggests that programmers are reluctant to adopt new languages to
-Security-oriented programming languages do not enjoy wide adoption.
+Web applications must ultimately generate strings that contain commands to be consumed by systems like web browsers and database engines.
+If these strings are constructed from user input that has not been properly sanitized, this can expose costly injection vulnerabilities.
+%Structured frameworks for constructing commands centralizes control over string sanitization, but verifying that the framework itself sanitizes user input correctly can still be tricky.
+
+In this paper, we introduce \emph{regular string types}, which classify strings known statically to be in a specified regular language.
+The language of a string is tracked by the type system through operations like concatenation, substitution and coercion, so regular string types can be used to implement, in essentially a conventional manner, the parts of a web application or application framework that constructs  commands.
+Simple type annotations at key points can be used to statically verify that sanitization has been performed correctly without introducing redundant run-time checks.
+
+We take the position that to be practical, such a type system cannot require that programmers adopt a new programming language.
+Ideally, such a ``special-purpose'' type system would be distributed as a library that could safely be used together with other such type systems.
+We support this by specifying a sound translation to a simple language containing only strings and regular expressions, then discuss implementing the type system together with this translation as a library in \texttt{atlang}, an extensible static type system for Python (being developed by the authors).% represents the most sophisticated application of this library to date.
+% Security-oriented programming languages do not enjoy wide adoption.
 Conversely, library and framework-based approaches to security are pervasive.
@@
 The base case ($r=\alpha, \alpha \in \Sigma$) is trivial.
-The choice and sequential cases require lemma 1, and prove a more general property that $\lsubst{r}{s_1}{s_2} \in \lang{r'}$ for some $r'$ equivalent to $\lreplace{r}{r_1}{r_2}$.
+The choice and sequential cases require lemma 1, and prove a more general property that $\lsubst{r}{s_1}{s_2} \in \lang{r'}$ for some $r'$ isomorphic to $\lreplace{r}{r_1}{r_2}$.
 The closure case follows from the induction hypothesis applied to any $n$-unwinding.

### 2014-08-06 sanitation-psp14/sanitation-psp14-short.tex
Fragment -> full system, but the rules don't fit on a single page so we might have to revert (or else get fancy with space)
Fixed typos due to out-of-date \dosubst macro
Factored s \in \lang{r} out of the theorem previously numbered as 10.
This property is now a property of \lambdas (thm 9) which is preserved under
translation (them 13). In order to prove them 13, we need thm 12 which is
basically CF + thm 11

--- 2014-08-05 sanitation-psp14/sanitation-psp14-short.tex
@@
 
+\newcommand{\val}{{\sf val}}
+
 \newcommand{\strin}[1]{\sistr{#1}}
@@
 % Meta-theoretic functions
-\newcommand{\dosubst}[3]{{\tt replace}(#1,#2,#3)}
 \newcommand{\lsubst}[3]{{\tt lsubst}(#1,#2,#3)} % This used to render lreplace(...) so there're probably mistkes wherever this command was used now.
@@
 
-<$\psi$> ::=	...			\hfill	source types					\alt
+<$\psi$> ::=	$\psi \rightarrow \psi$			\hfill	source types					\alt
 $\stringin{r}$
 
-<S> ::= ... \hfill source terms \alt
+<S> ::= 
+      $\lambda x . e$ \hfill source terms \alt
+      $e e$ \alt
       $\strin{s}$ \hfill $s \in \Sigma^{*}$ \alt
@@
 
-<$\theta$> ::= ... \hfill target types \alt
+<$\theta$> ::= $\theta \rightarrow \theta$ \hfill target types \alt
 $\str$ \alt $\regex$
 
-<P> ::= ... \hfill target terms \alt
+<P> ::= $\lambda x . e$ \hfill target terms \alt
+  $e e$ \alt
   $\tstr{s}$ \alt
@@
 % is modified before and after this figure.
+
 \addtocounter{figure}{4}
@@
 \begin{mathpar}
+%\inferrule[S-T-Abs]
+%{\sctx, x : T_1 \vdash e : T_2}
+%{\sctx \vdash \lambda x.e : T_1 \rightarrow T_2}
+%
+%\inferrule[S-T-App]
+%{\sctx \vdash e_1 : T' \\ \sctx \vdash e_2 : T' \rightarrow T}
+%{\sctx \vdash e_1 e_2 : T}
+%  
 \inferrule[S-T-Stringin-I]
@@
 
-\inferrule[S-T-replace]
-{\sctx \vdash S_1 : \stringin{r_1} \\ \sctx \vdash S_2 : \stringin{r_2} \\\\ {\tt lsubst}(r,r_1,r_2) = r'}
+\inferrule[S-T-Replace]
+{\sctx \vdash S_1 : \stringin{r_1} \\ \sctx \vdash S_2 : \stringin{r_2} \\\\ \lreplace{r}{r_1}{r_2} = r'}
 {\sctx \vdash \sisubst{r}{S_1}{S_2} : \stringin{r'}}
@@
 \begin{mathpar}
+%
+%\inferrule[S-E-Abs]                                                             
+%{ \ }                                                                           
+%{\lambda x.e \sreduces \lambda x.e}                                             
+%
+%\inferrule[S-E-App]
+%{ e_1 \sreduces e_1' \\ e_2 \sreduces e_2' }
+%{ e_1 e_2 \sreduces e_1' e_2'}
+%
 \inferrule[S-E-RStr]
@@
 \inferrule[S-E-Replace]
-{S_1 \sreduces \strin{s_1} \\ S_2 \sreduces \strin{s_2} \\ \dosubst{r}{s_1}{s_2} = s} 
+{S_1 \sreduces \strin{s_1} \\ S_2 \sreduces \strin{s_2} \\ \lsubst{r}{s_1}{s_2} = s} 
 {\sisubst{r}{S_1}{S_2} \sreduces \sistr{s}}
@@
 \begin{mathpar}
+%\inferrule[P-T-Abs]
+%{\tctx, x : T_1 \vdash e : T_2}
+%{\tctx \vdash \lambda x.e : T_1 \rightarrow T_2}
+%
+%\inferrule[P-T-App]
+%{\tctx \vdash e_1 : T \\ \tctx \vdash e_2 : T' \rightarrow T}
+%{\tctx \vdash e_1 e_2 : T}
+%
 \inferrule[P-T-String]
@@
 \begin{mathpar}
+
+%\inferrule[P-E-Abs]
+%{ \ }
+%{\lambda x.e \sreduces \lambda x.e}
+%
+%\inferrule[P-E-App]
+%{ e_1 \sreduces e_1' \\  e_2 \sreduces e_2' }
+%{ e_1 e_2 \sreduces e_1' e_2'}
+%
   \inferrule[P-E-Str]
@@
 \inferrule[P-E-Replace]
-{P_1 \treduces \rx{r} \\ P_2 \treduces \tstr{s_2} \\ P_3 \treduces \tstr{s_3} \\ \dosubst{r}{s_2}{s_3} = s} 
+{P_1 \treduces \rx{r} \\ P_2 \treduces \tstr{s_2} \\ P_3 \treduces \tstr{s_3} \\ \lsubst{r}{s_2}{s_3} = s} 
 {\preplace{P_1}{P_2}{P_3} \treduces \tstr{s}}
@@
 \begin{lem}{Properties of Regular Languages and Expressions.} \label{thm:regexprops}
-The following are well-known properties of regular expressions which are necessary for our proofs:
+The following are properties of regular expressions which are necessary for our proofs:
 If $s_1 \in \lang{r_1}$ and $s_2 \in \lang{r_2}$ then $s_1s_2 \in \lang{r_1r_2}$.
 For all strings $s$ and expressions $r$, either $s \in \lang{r}$ or $s \not \in \lang{r}$.
-Regular languages are closed under complements and concatenation.
-The regular expressions correspond bijectively to the regular languages.
 \end{lem}
@@
 \end{proof}
-
 \begin{lem}
@@
 
+In addition to safety, we proof a correctness result for $\lambdas$ which will
+be preserved under translation.
+
+\begin{thm}{Input Sanitation Correctness of $\lambdas$}\label{thm:scorrect}
+  If $\sctx \vdash S : \stringin{r}$ then either (a) $S \sreduces \rstr{s}$ for $s \in \lang{r}$
+  or (b) $S \err$.
+\end{thm}
+\begin{proof}
+  We may actually want a proof here.
+\end{proof}
+
+Finally, $\lambdap$ is also type-safe. We omit the proof due to space constraints,
+and because the proof is trivial.
+
 \begin{thm}
@@
 
-We now present the main correctness result.
-
-\begin{thm}
-If $S : \strin{r}$ then there exists a P such that $\trden{s} = P$ and either: 
-(a) $P \treduces \tstr{s}$ and $S \sreduces \strin{s}$, and $s \in lang{r}$; or
+\begin{thm}{Translation Correctness} \label{thm:tcorrect}
+If $\sctx \vdash S : \stringin{r}$ then there exists a $P$ such that $\trden{S} = P$ and either: 
+(a) $P \treduces \tstr{s}$ and $S \sreduces \strin{s}$, or
 (b) $P \err$ and $S \err$.
@@
 are necessary.
-
+\end{proof}
+
+\begin{thm}
+If If $\sctx \vdash S : \stringin{r}$ and $\trden{S} = P$ then either:
+(a) $P \treduces \tstr{s}$ and $S \sreduces \strin{s}$, or                      
+(b) $P \err$ and $S \err$. 
+\end{thm}
+\begin{proof}
+  The theorem  follows directly from canonical forms for translations and theorem \ref{thm:tcorrect}.
+\end{proof}
+
+Finally, our main theorem establishes that input sanitation correctness of $\lambdas$ is preserved under the translation into $\lambdap$.
+
+\begin{thm}{Input Sanitation Correctness for Translated Terms}
+  If $\trden{S} = P$ and $\sctx \vdash S : \stringin{r}$ then either $P \err$ or $P \sreduces \tstr{s}$
+  for $s \in \lang{r}$.
+\end{thm}
+\begin{proof}
+  By theorem \ref{tcorrect}, $P \sreduces \tstr{s}$ implies that $S \sreduces \strin{s}$.
+  By theorem \ref{scorrect}, this together with the assumption that $S$ is well-typed implies that $s \in \lang{r}$.
+\end{proof}
+% PROOF OF TRANSLATION CORRECTNESS.
+% This proof is slightly out of say. It's SUFFICIENT, but is actually more
+% than sufficient because we prove \ref{scorrect} along with translation
+% correctness....
 %\begin{description}
@@
 %\end{description}
-\end{proof}
 

### 2014-08-06 sanitation-psp14/sanitation-psp14-short.tex
Fixed some refs

@@
 
-\begin{thm}{Input Sanitation Correctness of $\lambdas$}\label{thm:scorrect}
+\begin{thm}{Input Sanitation Correctness of $\lambdas$.}\label{thm:scorrect}
   If $\sctx \vdash S : \stringin{r}$ then either (a) $S \sreduces \rstr{s}$ for $s \in \lang{r}$
@@
 \begin{proof}
-  We may actually want a proof here.
+  We may actually want to present this proof.
 \end{proof}
@@
 \begin{proof}
-  By theorem \ref{tcorrect}, $P \sreduces \tstr{s}$ implies that $S \sreduces \strin{s}$.
-  By theorem \ref{scorrect}, this together with the assumption that $S$ is well-typed implies that $s \in \lang{r}$.
+  By theorem \ref{thm:tcorrect}, $P \sreduces \tstr{s}$ implies that $S \sreduces \strin{s}$.
+  By theorem \ref{thm:scorrect}, this together with the assumption that $S$ is well-typed implies that $s \in \lang{r}$.
 \end{proof}

### 2014-08-06 sanitation-psp14/sanitation-psp14-short.tex
Cleaned up regex lemmas and replaced wall of text at the beginning of section 2 with a concise overview of the section. We probably still need some fill text, but it should be interspersed and informative.

@@
 
-The $\lambdas$ language is characterized by a type of strings indexed by regular
-expressions, together with operations on such strings which correspond to common
-input sanitation patterns.
-This section presents the grammar, typing rules and operational semantics for
-$\lambdas$ as well as an underlying language $\lambdap$.
-
-The system $\lambdas$ is the simply typed lambda calculus extended with \emph{regular expression types}, which are string
-types ensuring a string belongs to a specified language. For instance, 
-$S : \stringin{r}$ reads ``$s$ is a string matching $r$".
-the system includes an operation for replacing all instances of a pattern $r$ in a string $s_1$ with another string $s_2$. 
-Input sanitation algorithms -- as implemented by developers or within popular libraries and frameworks -- are often implemented in terms of this replace operation.
-
-The language $\lambdap$ is a simple functional language extended with a minimal regular expression library.
-Any general purpose programming language could stand in for $\lambdap$; for instance, SML or Python.
-In an implementation, our correctness results are modulo the underlying language's correct implementation of regular expression matching (see P-E-Replace).
-
-Finally, we define a translation from our type system $\lambdas$ into $\lambdap$ which preserves the safety guarnatee codified in the string types of $\lambdas$.
+In this section we define a language for statically checked string sanitation ($\lambdas$)
+and prove that its correctness property is preserved under translation to a language
+with regular expression matching capabilities ($\lambdap$). A brief outline of this section
+follows:
+\begin{itemize}
+  \item Page 3 contains a definition of $\lambdas, \lambdap$ and the translation from $\lambdas$ to $\lambdap$.
+  \item In \S 2.1 we state some properties about regular expressions which are needed in our correctness proofs.
+  \item In \S 2.2 we prove type safety for $\lambdap$ as well as both type safety and correctness for $\lambdas$.
+  \item In \S 2.3 we prove that translation preserves the correctness reesult about $\lambdas$.
+\end{itemize}
+
+%The $\lambdas$ language is characterized by a type of strings indexed by regular
+%expressions, together with operations on such strings which correspond to common
+%input sanitation patterns.
+%This section presents the grammar, typing rules and operational semantics for
+%$\lambdas$ as well as an underlying language $\lambdap$.
+%
+%The system $\lambdas$ is the simply typed lambda calculus extended with \emph{regular expression types}, which are string
+%types ensuring a string belongs to a specified language. For instance, 
+%$S : \stringin{r}$ reads ``$s$ is a string matching $r$".
+%the system includes an operation for replacing all instances of a pattern $r$ in a string $s_1$ with another string $s_2$. 
+%Input sanitation algorithms -- as implemented by developers or within popular libraries and frameworks -- are often implemented in terms of this replace operation.
+%
+%The language $\lambdap$ is a simple functional language extended with a minimal regular expression library.
+%Any general purpose programming language could stand in for $\lambdap$; for instance, SML or Python.
+%In an implementation, our correctness results are modulo the underlying language's correct implementation of regular expression matching (see P-E-Replace).
+%
+%Finally, we define a translation from our type system $\lambdas$ into $\lambdap$ which preserves the safety guarnatee codified in the string types of $\lambdas$.
 Because our extension is conservative and its guarnatees are preserved under translation to an underlying language, it is highly composable with other analyses.
-
-Unfortunately, we are unable to present full proofs in this paper due to space constraints; however, proofs are given in full in the accompanying technical report?.
-%Note that we never insert an internal check where the type of the string implies
-%that a check must succeed. Furthermore, expensive calculations (such as language
-%inclusion or ${\sf rreplace}$) all occur at compile time.
-%Since sanitation problems are generally expressible with smalle, simple expressions,
-%we believe that the compile-time overhead is not significant enough to prohibiour target usecase. Informal experimentation with our implementation seems to
-%support this assertion.
+%
+%Unfortunately, we are unable to present full proofs in this paper due to space constraints; however, proofs are given in full in the accompanying technical report?.
+%%Note that we never insert an internal check where the type of the string implies
+%%that a check must succeed. Furthermore, expensive calculations (such as language
+%%inclusion or ${\sf rreplace}$) all occur at compile time.
+%%Since sanitation problems are generally expressible with smalle, simple expressions,
+%%we believe that the compile-time overhead is not significant enough to prohibiour target usecase. Informal experimentation with our implementation seems to
+%%support this assertion.
 
@@
 \begin{defn}[$\tt{lsubst}$]
-The function $\lsubst{r}{s_1}{s_2}$ produces a string in which all substrings of $s_1$ matching $r$ are replaced with $s_2$.
+The replation $\lsubst{r}{s_1}{s_2} = s$ produces a string $s$ in which all substrings of $s_1$ matching $r$ are replaced with $s_2$.
 \end{defn}
@@
 \begin{defn}[$\tt{lreplace}$]
-The function $\lreplace{r}{r_1}{r_2}$ produces a regular expression in which any sublanguage $\lang{r_1'}$ of $\lang{r_1}$ satisfying the condition $\lang{r_1'} \subseteq{\lang{r}}$ is replaced with $\lang{r_2}$.
+  The relation $\lreplace{r}{r_1}{r_2} = r'$ relates $r, r_1,$ and $r_2$ to
+  a language $r'$ containing all strings of $r_1$ except that any substring $s_{pre} s s_{post} \in \lang{r_1}$ where $s \in \lang{r}$
+  is replaced by the set of strings $s_{pre} s_2 s_{post}$ for all $s_2 \in \lang{r_2}$ (the prefix and postfix positions may be empty).
 \end{defn}
 
-\begin{lem}{Closure and Totality of Replacement.} \label{thm:total}
-If $r, r_1$ and $r_2$ are regular expressions, then $\lreplace{r}{r_1}{r_2}$ is also a regular expression.
+\begin{lem}{Closure.} \label{thm:total}
+  If $\lang{r}, \lang{r_1}$ and $\lang{r_2}$ are regular expressions, then $\lang{\lreplace{r}{r_1}{r_2}}$ is also a regular language.
 \end{lem}
 \begin{proof}
-By induction on $r$ and closure properties of regular expressions.
+The theorem follows from closure under difference, right quotient and reversal.
 \end{proof}
@@
 \begin{lem}{Substitution Correspondence.} \label{thm:substcorrespondence}
-If $s_1 \in \lang{r_1}$ and $s_2 \in \lang{r_2}$ then $\lsubst{r}{s_1}{s_2} \in \lang{\lreplace{r}{r_1}{r_2}}$.
+  If $s_1 \in \lang{r_1}$ and $s_2 \in \lang{r_2}$ then $\lsubst{r}{s_1}{s_2} \in \lang{\lreplace{r}{s_1}{s_2}}$.
 \end{lem}
 \begin{proof}
-The proof proceeds by structural induction on $r$.
-The base case ($r=\alpha, \alpha \in \Sigma$) is trivial.
-The choice and sequential cases require lemma 1, and prove a more general property that $\lsubst{r}{s_1}{s_2} \in \lang{r'}$ for some $r'$ equivalent to $\lreplace{r}{r_1}{r_2}$.
-The closure case follows from the induction hypothesis applied to any $n$-unwinding.
+The theorem follows from the refinitions of lsubst and lreplace; note that language substitutions over-approximate string substititons.
+\end{proof}
 
@@
 
-\begin{thm}{Safety for the String Fragment of P.}
+\begin{thm}{Safety and Sanitation Correctness for the String Fragment of P.} \label{thm:scorrect}
   Let $S$ be a term in the source language. If $\sctx \vdash S : \stringin{r}$ 
-  then $S \sreduces \rstr{s}$ and $\rstr{s} : \stringin{r}$, or else $S \err$.
+  then $S \sreduces \rstr{s}$, $\sctx \rstr{s} : \stringin{r}$, and $s \in \lang{r}$; or else $S \err$.
 \end{thm}
 \begin{proof}
-By induction on the derivation of $\sctx \vdash S : \psi$. The interesting case
-is S-T-Replace, which requires lemma \ref{thm:substcorrespondence}. The 
-Coercion lemma additionally requires the second property of lemma \ref{thm:regexprops}.
-
-%\begin{description}
-%\item[S-T-Stringin-I:]
-%If $S = \rsconcat{S_1}{S_2} : \stringin{r}$ then
-%$S \sreduces S$ by S-E-RStr, and $\sctx \vdash S : \psi$ by assumption.
-%
-%\item[S-T-Concat:] 
-%Suppose $S = \rsconcat{S_1}{S_2} : \stringin{r_1 r_2}$.
-%By inversion, $\sctx \vdash S_1 : \stringin{r_1}$ and $\sctx \vdash S_1 : \stringin{r_2}$.
-%It follows by induction that either $S_1 \err$, $S_2 \err$, 
-%or $S_1 \sreduces \rstr{s_1}$ and $S_2 \sreduces \rstr{s_2}$ for some $s_1 \in \lang{r_1}$ and $s_2 \in \lang{r_2}$.
-%In the latter case $S \reduces \rstr{s_1s_2}$ by S-E-Concat and $\sctx \vdash \rstr{s_1s_2} : \tstr{r_1r_2}$ by \ref{thm:regexprops}.
-%In the former cases, $S \err$.
-%
-%\item[S-T-Replace:a]
-%Suppose $S = \rreplace{r}{S_1}{S_2}$ and $\sctx \vdash S : \stringin{r'}$.
-%By inversion $\sctx \vdash S_1 : \stringin{r_1}$ and $\sctx \vdash S_2 : \stringin{r_2}$ such that $\lsubst{r}{r_1}{r_2} = r'$.
-%By induction, $S_1 \err$, $S_2 \err$ 
-%or $S_1 \sreduces \rstr{s_1}$ and $S_2 \sreduces \rstr{s_2}$ such that 
-%In the latter case, we know $\lreplace{r}{s_1}{s_2} \in
-%\lang{\lsubst{r}{r_1}{r_2}}$ by lemma \ref{thm:substcorrespondence};
-%therefore by S-E-Replace, $S \sreduces \rstr{s}$ such that  
-%
-%$s \in \lang{\lsubst{r}{r_1}{r_2}} = \lang{r'}$.
-%So by S-T-String-I, $\rstr{s} : \stringin{r'}$.
-%In the former cases, $S \err$.
-%
-%\item[S-T-Coerce:]
-%Suppose $S = \coerce{r}{S_1}$ and $S : \stringin{r}$.
-%By inversion, $\sctx \vdash S_1 : \stringin{r'}$.
-%By induction, $S_1 \err$ or $S_1 \sreduces \rstr{s}$.
-%In the former case $S \err$ by propagation rules.
-%In the latter case we have by property 2 of \ref{thm:regexprops} that $s \in \lang{r}$ or else
-%$s \not \in \lang{r}$.
-%If $s \in \lang{r}$ then $\rstr{s} : \stringin{r}$.
-%If $s \not \in \lang{r}$ then $S \err$. 
-%\end{description}
+By induction on the typing relation, where (a) case holds by lemma \ref{thm:regexprops} in the S-T-Concat case and lemma \ref{thm:substcorrespondence} in the S-T-Replace case..
+The (b) cases hold by unstated, but standard, error propagation rules.
 \end{proof}
@@
 
-\begin{thm}{Input Sanitation Correctness of $\lambdas$.}\label{thm:scorrect}
-  If $\sctx \vdash S : \stringin{r}$ then either (a) $S \sreduces \rstr{s}$ for $s \in \lang{r}$
-  or (b) $S \err$.
-\end{thm}
-\begin{proof}
-  We may actually want to present this proof.
-\end{proof}
-
-Finally, $\lambdap$ is also type-safe. We omit the proof due to space constraints,
-and because the proof is trivial.
+%\begin{thm}{Correctness of Input Sanitation (CIS) for $\lambdas$.}\label{thm:scorrect}
+%  If $\sctx \vdash S : \stringin{r}$ then either (a) $S \sreduces \rstr{s}$ for $s \in \lang{r}$
+%  or (b) $S \err$.
+%\end{thm}
 
@@
   Let $P$ be a term in the target language. If $\tctx \vdash P : \theta$ 
-  then $P \sreduces P'$ and $\tctx \vdash P' : \theta$.
+  then $P \sreduces P'$ and $\tctx \vdash P' : \theta$, or else $P \err$.
 \end{thm}
@@
 
-\begin{thm}{Input Sanitation Correctness for Translated Terms}
+\begin{thm}{Correctness of Input Sanitation for Translated Terms.}
   If $\trden{S} = P$ and $\sctx \vdash S : \stringin{r}$ then either $P \err$ or $P \sreduces \tstr{s}$

### 2014-08-06 sanitation-psp14/sanitation-psp14-short.tex
Cleaned up safety/correctness divide

@@
   Let $S$ be a term in the source language. If $\sctx \vdash S : \stringin{r}$ 
-  then $S \sreduces \rstr{s}$, $\sctx \rstr{s} : \stringin{r}$, and $s \in \lang{r}$; or else $S \err$.
+  then $S \sreduces \rstr{s}$ and $\sctx \vdash \rstr{s} : \stringin{r}$; or else $S \err$.
 \end{thm}
@@
 
+\begin{thm}{Correctness of Input SAnitation for $\lambdas$.}\label{thm:scorrect}
+  If  $\sctx \vdash S : \stringin{r}$ and $S \sreduces \rstr{s}$ then $s \in \lang{r}$.
+\end{thm}
+\begin{proof}
+  Follows directly from type safety, canonical forms for $\lambdas$.
+\end{proof}
+
 %\begin{thm}{Correctness of Input Sanitation (CIS) for $\lambdas$.}\label{thm:scorrect}
@@
 
-\begin{thm}{Translation Correctness} \label{thm:tcorrect}
+\begin{thm}{Translation Correctness.} \label{thm:tcorrect}
 If $\sctx \vdash S : \stringin{r}$ then there exists a $P$ such that $\trden{S} = P$ and either: 
@@
 are necessary.
-\end{proof}
-
-\begin{thm}
-If If $\sctx \vdash S : \stringin{r}$ and $\trden{S} = P$ then either:
-(a) $P \treduces \tstr{s}$ and $S \sreduces \strin{s}$, or                      
-(b) $P \err$ and $S \err$. 
-\end{thm}
-\begin{proof}
-  The theorem  follows directly from canonical forms for translations and theorem \ref{thm:tcorrect}.
 \end{proof}

### 2014-08-06 sanitation-psp14/sanitation-psp14-short.tex
Changes to soft stuff

@@
 \section{Introduction}\label{intro}
-Improper input sanitation is a leading cause of security vulnerabilities in web applications [OWASP].
+Improper input sanitation is a leading cause of security vulnerabilities in web applications \cite{OWASP}.
 Command injection attacks exploit improper input sanitation by inserting malicious code into an otherwise benign command.
@@
 
-A number of research languages provide static guarantees that a program is free of input sanitation vulnerabilities [Jif][Ur/Web]. 
+A number of research languages provide static guarantees that a program is free of input sanitation vulnerabilities \cite{UrFlowOSDI10}. 
 Unlike this work, our solution to the input sanitation problem has a very low barrier to adoption; for instance, our implementation conservatively extends Python -- a popular language among web developers.
@@
 
-We are also unaware of any extensible programming languages which emphasize applications to security concerns (TRUE?).
+We are also unaware of any extensible programming languages which emphasize applications to security concerns.
 
 Incorporating regular expressions into the type system is not novel.
-The XDuce system \cite{pierce} typechecks XML schemas using regular expressions.
+The XDuce system \cite{HosoyaVouillonPierce2000ICFP,HosoyaPierce2002} checks XML documents against schemas using regular expressions.
+Similarly, XHaskell \cite{xhaskell} focuses on XML documents.
 We differ from this and related work in at least two ways.
@@
 
-In this section we define a language for statically checked string sanitation ($\lambdas$)
-and prove that its correctness property is preserved under translation to a language
-with regular expression matching capabilities ($\lambdap$). A brief outline of this section
-follows:
+In this section we define a language for statically checked string sanitation ($\lambdas$).
+The system has regular expression types $\stringin{r}$ whre $r$ is a regular expression. 
+Expressions of this type evaluate to string literals in the language described by $r$.
+Operations on expressions of type $\stringin{r}$ preserve this property; for instance,
+the result of string concatenation is
+
+The premier operation for manipulating strings in $\lambdas$ is string substitution, 
+which is a familiar operation to any programmer who has used regular expressions.
+The replacement operation replaces all instances of a pattern in one string with
+another string; for instance, $\lsubst(a|b, ``a", ``"c") = c$.
+In order to computer the type resulting from aubstitution, we also need to compute
+the result of replacing on language with another inside a given language.
+Finally, just for convienance, we provide a coerce operation. The introduction of
+coercion requires handling of runtime errors.
+
+The underlying language $\lambdap$ has only one type for strings. We prove that whenever
+a term is translated from $\lambdas$ to $\lambdap$, correctness is preserved.
+The only exception is in the case of unsafe casts in $\lambdas$, which are unnecessary but
+are included to demonstrate that the regex library of $\lambdap$ may be used to insert
+dynamic checks whenever even when developers are not careful about using statically checked
+operations.
+
+A brief outline of this section follows:
 \begin{itemize}
@@
 regular expressions $r$ are defined. throughout the paper, $\lang{r}$ refers to the
-language recognized by the expression $r$. This distinction between the expression
+language recognized by the regular expression $r$. This distinction between the regular expression
 and its language -- typically elided in the literature -- makes our definition and
@@
 If $s_1 \in \lang{r_1}$ and $s_2 \in \lang{r_2}$ then $s_1s_2 \in \lang{r_1r_2}$.
-For all strings $s$ and expressions $r$, either $s \in \lang{r}$ or $s \not \in \lang{r}$.
+For all strings $s$ and regular expressions $r$, either $s \in \lang{r}$ or $s \not \in \lang{r}$.
+Regular languages are closed under difference, right quotient, reversal, and string homomorphism.
 \end{lem}
+
+If any of these properties are unfamiliar, the reader may refer to a standard text on the subject \cite{cinderella}.
 
@@
 \begin{proof}
-The theorem follows from closure under difference, right quotient and reversal.
+The theorem follows from closure under difference, right quotient, reversal and string homomorphism.
 \end{proof}
@@
 \begin{proof}
-The theorem follows from the refinitions of lsubst and lreplace; note that language substitutions over-approximate string substititons.
+The theorem follows from the definitions of lsubst and lreplace; note that language substitutions over-approximate string substititons.
 \end{proof}
@@
 
-In addition to safety, we proof a correctness result for $\lambdas$ which will
-be preserved under translation.
-
-\begin{thm}{Correctness of Input SAnitation for $\lambdas$.}\label{thm:scorrect}
+In addition to safety, we proof a correctness result for $\lambdas$ which ensures that well-typed terms of regular string type are in the language associated with their type.
+
+\begin{thm}{Correctness of Input Sanitation for $\lambdas$.}\label{thm:scorrect}
   If  $\sctx \vdash S : \stringin{r}$ and $S \sreduces \rstr{s}$ then $s \in \lang{r}$.
@@
 
+\section{Implementation}
+
+An implementation of a similar system as a type system extension in the programming language Ace is discussed in \cite{fulton12} and \cite{fulton13}.
+The implementation only supports a special case of replacement where unsafe substrings are simply stripped.
+In practice, most libraries and frameworks escape command sequences instead of stripping characters; therefore, we plan to extend our implementation with
+support for the replace operation as described in this paper..
+
 \section{Conclusion}
@@
 Unlike other approaches, our solution complements existing, familiar solutions while providing a strong guarantee that traditional library and framework-based approaches are implemented and utilized correctly.
-
-Papers that needs to be cited in this section:
-\begin{itemize}
-  \item Ur/Web OSDI paper
-  \item Jif?
-  \item OWASP
-  \item XDuce and related papers. 
-  \item src?
-  \item Ace or Wyvern paper?
-  \item hotsos?
-  \item Haskell extension paper
-  \item Maybe some popular FOSS libraries/frameworks that do input sanitation?
-\end{itemize}
 
@@
 %
-%\bibliographystyle{abbrv}
-%
-%% The bibliography should be embedded for final submission.
-%
-%\bibliography{../research}
-%%\softraggedright
-%%P. Q. Smith, and X. Y. Jones. ...reference text...
-%
-%
-
+\bibliographystyle{abbrv}
+
+\bibliography{research}
+
+% The bibliography should be embedded for final submission.
+
+%\begin{thebibliography}
+%
+%\bibitem{lamport94}
+%Leslie Lamport,
+%\emph{\LaTeX: a document preparation system}.
+%Addison Wesley, Massachusetts,
+%2nd edition,
+%1994.
+%
+%\end{thebibliography}
 \end{document}

### 2014-08-06 sanitation-psp14/sanitation-psp14-short.tex
Updated out-of-date intro text.

@@
 Similarly, XHaskell \cite{xhaskell} focuses on XML documents.
-We differ from this and related work in at least two ways.
-First, our system  is defined within an extensible type system; second, and more importantly, we have demonstrated that regular expression types are applicable to the web security domain.
-
-In conclusion, our system is novel in at least two ways:
+We differ from this and related work in at least three ways:
 \begin{itemize}
-\item The safety guarantees provided by libraries and frameworks in popular languages
-are not as (statically) justified as is often belived (or even claimed).
-\item Our extension is the first major demonstration of how an extensible type
-system may be used to provide light-weight, composable security analyses based upon
-idiomatic code.
+  \item Our system is defined within an extensible type system.
+  \item We demonstrate that regular expression types are applicable to the web security domain, whereas previous work on regular expression types focused on XML schemas.
+  \item Although our static replacement operation is definable in some languages with regular expression types, we are the first to expose this operation and connect the semantics of regular language replacement with the semantics of string substitution via a type safety and compilation correctness argument.
 \end{itemize}
+
+In conclusion, our contribution is a type system, implemented within an extensible type system, for checking the correctness of input sanitation algorithms.
 
@@
 Expressions of this type evaluate to string literals in the language described by $r$.
-Operations on expressions of type $\stringin{r}$ preserve this property; for instance,
-the result of string concatenation is
+Operations on expressions of type $\stringin{r}$ preserve this property.
 
@@
 The replacement operation replaces all instances of a pattern in one string with
-another string; for instance, $\lsubst(a|b, ``a", ``"c") = c$.
+another string; for instance, $\lsubst{a|b}{a}{c} = c$.
 In order to computer the type resulting from aubstitution, we also need to compute

### 2014-08-06 sanitation-psp14/sanitation-psp14-short.tex
typos

@@
 Doing so requires reasoning about the operations on regular languages corresponding to standard operations on strings; we are unaware of any production system which contains this form of reasoning.
-Therefore, even where frameworks and libraries provide a viable interface or wrapper around input sanitation, our approach is complementary because it ensurees the correctness of the framework or library itself.
+Therefore, even where frameworks and libraries provide a viable interface or wrapper around input sanitation, our approach is complementary because it ensures the correctness of the framework or library itself.
 Furthermore, our approach is more general than database abstraction layers because our mechanism is applicable to all forms of command injection (e.g.

### 2014-08-07 sanitation-psp14/sanitation-psp14-short.tex
changes

--- 2014-08-06 sanitation-psp14/sanitation-psp14-short.tex
@@
 
-\newcommand{\val}{{\sf val}}
-
 \newcommand{\strin}[1]{\sistr{#1}}
@@
 % Meta-theoretic functions
+\newcommand{\dosubst}[3]{{\tt replace}(#1,#2,#3)}
 \newcommand{\lsubst}[3]{{\tt lsubst}(#1,#2,#3)} % This used to render lreplace(...) so there're probably mistkes wherever this command was used now.
@@
 %Evidence suggests that programmers are reluctant to adopt new languages to
-Web applications must ultimately generate strings that contain commands to be consumed by systems like web browsers and database engines.
-If these strings are constructed from user input that has not been properly sanitized, this can expose costly injection vulnerabilities.
+Web applications must ultimately generate strings containing commands that are   executed by external systems like web browsers and database engines.
+If these strings are constructed from user input that has not been properly sanitized, this can lead to a variety of command injection vulnerabilities.
 %Structured frameworks for constructing commands centralizes control over string sanitization, but verifying that the framework itself sanitizes user input correctly can still be tricky.
 
-In this paper, we introduce \emph{regular string types}, which classify strings known statically to be in a specified regular language.
-The language of a string is tracked by the type system through operations like concatenation, substitution and coercion, so regular string types can be used to implement, in essentially a conventional manner, the parts of a web application or application framework that constructs  commands.
-Simple type annotations at key points can be used to statically verify that sanitization has been performed correctly without introducing redundant run-time checks.
-
-We take the position that to be practical, such a type system cannot require that programmers adopt a new programming language.
-Ideally, such a ``special-purpose'' type system would be distributed as a library that could safely be used together with other such type systems.
-We support this by specifying a sound translation to a simple language containing only strings and regular expressions, then discuss implementing the type system together with this translation as a library in \texttt{atlang}, an extensible static type system for Python (being developed by the authors).% represents the most sophisticated application of this library to date.
+In this paper, we introduce \emph{regular string types}, which classify strings known statically to be in a specified regular language and support operations like concatenation, substitution and coercion.
+Regular string types can be used to implement, in essentially a conventional manner, the parts of a web application or application framework that construct such command strings.
+Straightforward type annotations at key interfaces can be used to statically verify that sanitization has been performed correctly without introducing redundant run-time checks.
+We specify this type system as a minimal typed lambda calculus, $\lambda_{RS}$.
+
+To be practical, adopting a type system like this should not require adopting a new programming language.
+Instead, we favor extensible type systems: new static type systems like this should be distributed as libraries atop a mechanism that guarantees that they can be safely composed.
+We support this by 1) specifying a  translation from $\lambda_{RS}$ to a language containing only strings and regular expressions, then, taking Python as such a language, 2) implement the type system together with the translation as a library using \texttt{atlang}, an extensible static type system for Python (being developed by the authors).% represents the most sophisticated application of this library to date.
 % Security-oriented programming languages do not enjoy wide adoption.
@@
 \section{Introduction}\label{intro}
-Improper input sanitation is a leading cause of security vulnerabilities in web applications \cite{OWASP}.
+Improper input sanitation is a leading cause of security vulnerabilities in web applications [OWASP].
 Command injection attacks exploit improper input sanitation by inserting malicious code into an otherwise benign command.
@@
 When these methods are unavailable or insufficient, developers implement custom sanitation techniques.
-In both cases, sanitation algorithms are implemented using the language's regular expression capabilities and usually \emph{replace} potentially unsafe strings with equivalent escaped strings.
+Im both cases, sanitation algorithms are implemented using the language's regular expression capabilities and usually \emph{replace} potentially unsafe strings with equivalent escaped strings.
 
@@
 Libraries, frameworks or functions which construct and execute commands containing input can specify a safe subset $\stringin{r_{\sf spec}}$ of strings, and input sanitation algorithms can construct such a string using ${\sf rreplace}$ or, optionally, by coercion (in which case a runtime check is inserted).
-We also show how this system is translated into a host language containing a regular expression library such that the safety guarantee of the extended language is preserved.
+We also show how this simple can be translated into a host language containing a regular expression library such that the safety guarantee of the extended language is preserved.
 
@@
 
-A number of research languages provide static guarantees that a program is free of input sanitation vulnerabilities \cite{UrFlowOSDI10}. 
+A number of research languages provide static guarantees that a program is free of input sanitation vulnerabilities [Jif][Ur/Web]. 
 Unlike this work, our solution to the input sanitation problem has a very low barrier to adoption; for instance, our implementation conservatively extends Python -- a popular language among web developers.
@@
 
-We are also unaware of any extensible programming languages which emphasize applications to security concerns.
+We are also unaware of any extensible programming languages which emphasize applications to security concerns (TRUE?).
 
 Incorporating regular expressions into the type system is not novel.
-The XDuce system \cite{HosoyaVouillonPierce2000ICFP,HosoyaPierce2002} checks XML documents against schemas using regular expressions.
-Similarly, XHaskell \cite{xhaskell} focuses on XML documents.
+The XDuce system \cite{pierce} typechecks XML schemas using regular expressions.
 We differ from this and related work in at least two ways.
@@
 
-In this section we define a language for statically checked string sanitation ($\lambdas$).
-The system has regular expression types $\stringin{r}$ whre $r$ is a regular expression. 
-Expressions of this type evaluate to string literals in the language described by $r$.
-Operations on expressions of type $\stringin{r}$ preserve this property; for instance,
-the result of string concatenation is
-
-The premier operation for manipulating strings in $\lambdas$ is string substitution, 
-which is a familiar operation to any programmer who has used regular expressions.
-The replacement operation replaces all instances of a pattern in one string with
-another string; for instance, $\lsubst(a|b, ``a", ``"c") = c$.
-In order to computer the type resulting from aubstitution, we also need to compute
-the result of replacing on language with another inside a given language.
-Finally, just for convienance, we provide a coerce operation. The introduction of
-coercion requires handling of runtime errors.
-
-The underlying language $\lambdap$ has only one type for strings. We prove that whenever
-a term is translated from $\lambdas$ to $\lambdap$, correctness is preserved.
-The only exception is in the case of unsafe casts in $\lambdas$, which are unnecessary but
-are included to demonstrate that the regex library of $\lambdap$ may be used to insert
-dynamic checks whenever even when developers are not careful about using statically checked
-operations.
-
-A brief outline of this section follows:
-\begin{itemize}
-  \item Page 3 contains a definition of $\lambdas, \lambdap$ and the translation from $\lambdas$ to $\lambdap$.
-  \item In \S 2.1 we state some properties about regular expressions which are needed in our correctness proofs.
-   \item In \S 2.2 we prove type safety for $\lambdap$ as well as both type safety and correctness for $\lambdas$.
-  \item In \S 2.3 we prove that translation preserves the correctness reesult about $\lambdas$.
-\end{itemize}
-
-%The $\lambdas$ language is characterized by a type of strings indexed by regular
-%expressions, together with operations on such strings which correspond to common
-%input sanitation patterns.
-%This section presents the grammar, typing rules and operational semantics for
-%$\lambdas$ as well as an underlying language $\lambdap$.
-%
-%The system $\lambdas$ is the simply typed lambda calculus extended with \emph{regular expression types}, which are string
-%types ensuring a string belongs to a specified language. For instance, 
-%$S : \stringin{r}$ reads ``$s$ is a string matching $r$".
-%the system includes an operation for replacing all instances of a pattern $r$ in a string $s_1$ with another string $s_2$. 
-%Input sanitation algorithms -- as implemented by developers or within popular libraries and frameworks -- are often implemented in terms of this replace operation.
-%
-%The language $\lambdap$ is a simple functional language extended with a minimal regular expression library.
-%Any general purpose programming language could stand in for $\lambdap$; for instance, SML or Python.
-%In an implementation, our correctness results are modulo the underlying language's correct implementation of regular expression matching (see P-E-Replace).
-%
-%Finally, we define a translation from our type system $\lambdas$ into $\lambdap$ which preserves the safety guarnatee codified in the string types of $\lambdas$.
+The $\lambdas$ language is characterized by a type of strings indexed by regular
+expressions, together with operations on such strings which correspond to common
+input sanitation patterns.
+This section presents the grammar, typing rules and operational semantics for
+$\lambdas$ as well as an underlying language $\lambdap$.
+
+The system $\lambdas$ is the simply typed lambda calculus extended with \emph{regular expression types}, which are string
+types ensuring a string belongs to a specified language. For instance, 
+$S : \stringin{r}$ reads ``$s$ is a string matching $r$".
+the system includes an operation for replacing all instances of a pattern $r$ in a string $s_1$ with another string $s_2$. 
+Input sanitation algorithms -- as implemented by developers or within popular libraries and frameworks -- are often implemented in terms of this replace operation.
+
+The language $\lambdap$ is a simple functional language extended with a minimal regular expression library.
+Any general purpose programming language could stand in for $\lambdap$; for instance, SML or Python.
+In an implementation, our correctness results are modulo the underlying language's correct implementation of regular expression matching (see P-E-Replace).
+
+Finally, we define a translation from our type system $\lambdas$ into $\lambdap$ which preserves the safety guarnatee codified in the string types of $\lambdas$.
 Because our extension is conservative and its guarnatees are preserved under translation to an underlying language, it is highly composable with other analyses.
-%
-%Unfortunately, we are unable to present full proofs in this paper due to space constraints; however, proofs are given in full in the accompanying technical report?.
-%%Note that we never insert an internal check where the type of the string implies
-%%that a check must succeed. Furthermore, expensive calculations (such as language
-%%inclusion or ${\sf rreplace}$) all occur at compile time.
-%%Since sanitation problems are generally expressible with smalle, simple expressions,
-%%we believe that the compile-time overhead is not significant enough to prohibiour target usecase. Informal experimentation with our implementation seems to
-%%support this assertion.
+
+Unfortunately, we are unable to present full proofs here due to space constraints.
+%Note that we never insert an internal check where the type of the string implies
+%that a check must succeed. Furthermore, expensive calculations (such as language
+%inclusion or ${\sf rreplace}$) all occur at compile time.
+%Since sanitation problems are generally expressible with smalle, simple expressions,
+%we believe that the compile-time overhead is not significant enough to prohibiour target usecase. Informal experimentation with our implementation seems to
+%support this assertion.
 
@@
 
-\begin{figure}[h]
+\begin{figure}[t]
 \begin{grammar}
@@
 
-<$\psi$> ::=	$\psi \rightarrow \psi$			\hfill	source types					\alt
+<$\psi$> ::=	...			\hfill	source types					\alt
 $\stringin{r}$
 
-<S> ::= 
-      $\lambda x . e$ \hfill source terms \alt
-      $e e$ \alt
+<S> ::= ... \hfill source terms \alt
       $\strin{s}$ \hfill $s \in \Sigma^{*}$ \alt
@@
 
-<$\theta$> ::= $\theta \rightarrow \theta$ \hfill target types \alt
+<$\theta$> ::= ... \hfill target types \alt
 $\str$ \alt $\regex$
 
-<P> ::= $\lambda x . e$ \hfill target terms \alt
-  $e e$ \alt
+<P> ::= ... \hfill target terms \alt
   $\tstr{s}$ \alt
@@
 % is modified before and after this figure.
-
 \addtocounter{figure}{4}
@@
 \begin{mathpar}
-%\inferrule[S-T-Abs]
-%{\sctx, x : T_1 \vdash e : T_2}
-%{\sctx \vdash \lambda x.e : T_1 \rightarrow T_2}
-%
-%\inferrule[S-T-App]
-%{\sctx \vdash e_1 : T' \\ \sctx \vdash e_2 : T' \rightarrow T}
-%{\sctx \vdash e_1 e_2 : T}
-%  
 \inferrule[S-T-Stringin-I]
@@
 
-\inferrule[S-T-Replace]
-{\sctx \vdash S_1 : \stringin{r_1} \\ \sctx \vdash S_2 : \stringin{r_2} \\\\ \lreplace{r}{r_1}{r_2} = r'}
+\inferrule[S-T-replace]
+{\sctx \vdash S_1 : \stringin{r_1} \\ \sctx \vdash S_2 : \stringin{r_2} \\\\ {\tt lsubst}(r,r_1,r_2) = r'}
 {\sctx \vdash \sisubst{r}{S_1}{S_2} : \stringin{r'}}
@@
 \begin{mathpar}
-%
-%\inferrule[S-E-Abs]                                                             
-%{ \ }                                                                           
-%{\lambda x.e \sreduces \lambda x.e}                                             
-%
-%\inferrule[S-E-App]
-%{ e_1 \sreduces e_1' \\ e_2 \sreduces e_2' }
-%{ e_1 e_2 \sreduces e_1' e_2'}
-%
 \inferrule[S-E-RStr]
@@
 \inferrule[S-E-Replace]
-{S_1 \sreduces \strin{s_1} \\ S_2 \sreduces \strin{s_2} \\ \lsubst{r}{s_1}{s_2} = s} 
+{S_1 \sreduces \strin{s_1} \\ S_2 \sreduces \strin{s_2} \\ \dosubst{r}{s_1}{s_2} = s} 
 {\sisubst{r}{S_1}{S_2} \sreduces \sistr{s}}
@@
 \begin{mathpar}
-%\inferrule[P-T-Abs]
-%{\tctx, x : T_1 \vdash e : T_2}
-%{\tctx \vdash \lambda x.e : T_1 \rightarrow T_2}
-%
-%\inferrule[P-T-App]
-%{\tctx \vdash e_1 : T \\ \tctx \vdash e_2 : T' \rightarrow T}
-%{\tctx \vdash e_1 e_2 : T}
-%
 \inferrule[P-T-String]
@@
 \begin{mathpar}
-
-%\inferrule[P-E-Abs]
-%{ \ }
-%{\lambda x.e \sreduces \lambda x.e}
-%
-%\inferrule[P-E-App]
-%{ e_1 \sreduces e_1' \\  e_2 \sreduces e_2' }
-%{ e_1 e_2 \sreduces e_1' e_2'}
-%
 \inferrule[P-E-Str]
@@
 \inferrule[P-E-Replace]
-{P_1 \treduces \rx{r} \\ P_2 \treduces \tstr{s_2} \\ P_3 \treduces \tstr{s_3} \\ \lsubst{r}{s_2}{s_3} = s} 
+{P_1 \treduces \rx{r} \\ P_2 \treduces \tstr{s_2} \\ P_3 \treduces \tstr{s_3} \\ \dosubst{r}{s_2}{s_3} = s} 
 {\preplace{P_1}{P_2}{P_3} \treduces \tstr{s}}
@@
 regular expressions $r$ are defined. throughout the paper, $\lang{r}$ refers to the
-language recognized by the regular expression $r$. This distinction between the regular expression
+language recognized by the expression $r$. This distinction between the expression
 and its language -- typically elided in the literature -- makes our definition and
@@
 \begin{lem}{Properties of Regular Languages and Expressions.} \label{thm:regexprops}
-The following are properties of regular expressions which are necessary for our proofs:
+The following are well-known properties of regular expressions which are necessary for our proofs:
 If $s_1 \in \lang{r_1}$ and $s_2 \in \lang{r_2}$ then $s_1s_2 \in \lang{r_1r_2}$.
-For all strings $s$ and regular expressions $r$, either $s \in \lang{r}$ or $s \not \in \lang{r}$.
-Regular languages are closed under difference, right quotient, reversal, and string homomorphism.
+For all strings $s$ and expressions $r$, either $s \in \lang{r}$ or $s \not \in \lang{r}$.
+Regular languages are closed under complements and concatenation.
+The regular expressions correspond bijectively to the regular languages.
 \end{lem}
 
-If any of these properties are unfamiliar, the reader may refer to a standard text on the subject \cite{cinderella}.
-
 \begin{defn}[$\tt{lsubst}$]
-The replation $\lsubst{r}{s_1}{s_2} = s$ produces a string $s$ in which all substrings of $s_1$ matching $r$ are replaced with $s_2$.
+The function $\lsubst{r}{s_1}{s_2}$ produces a string in which all substrings of $s_1$ matching $r$ are replaced with $s_2$.
 \end{defn}
@@
 \begin{defn}[$\tt{lreplace}$]
-  The relation $\lreplace{r}{r_1}{r_2} = r'$ relates $r, r_1,$ and $r_2$ to
-  a language $r'$ containing all strings of $r_1$ except that any substring $s_{pre} s s_{post} \in \lang{r_1}$ where $s \in \lang{r}$
-  is replaced by the set of strings $s_{pre} s_2 s_{post}$ for all $s_2 \in \lang{r_2}$ (the prefix and postfix positions may be empty).
+The function $\lreplace{r}{r_1}{r_2}$ produces a regular expression in which any sublanguage $\lang{r_1'}$ of $\lang{r_1}$ satisfying the condition $\lang{r_1'} \subseteq{\lang{r}}$ is replaced with $\lang{r_2}$.
 \end{defn}
 
-\begin{lem}{Closure.} \label{thm:total}
-  If $\lang{r}, \lang{r_1}$ and $\lang{r_2}$ are regular expressions, then $\lang{\lreplace{r}{r_1}{r_2}}$ is also a regular language.
+\begin{lem}{Closure and Totality of Replacement.} \label{thm:total}
+If $r, r_1$ and $r_2$ are regular expressions, then $\lreplace{r}{r_1}{r_2}$ is also a regular expression.
 \end{lem}
 \begin{proof}
-The theorem follows from closure under difference, right quotient, reversal and string homomorphism.
+By induction on $r$ and closure properties of regular expressions.
 \end{proof}
@@
 \begin{lem}{Substitution Correspondence.} \label{thm:substcorrespondence}
-  If $s_1 \in \lang{r_1}$ and $s_2 \in \lang{r_2}$ then $\lsubst{r}{s_1}{s_2} \in \lang{\lreplace{r}{s_1}{s_2}}$.
+If $s_1 \in \lang{r_1}$ and $s_2 \in \lang{r_2}$ then $\lsubst{r}{s_1}{s_2} \in \lang{\lreplace{r}{r_1}{r_2}}$.
 \end{lem}
 \begin{proof}
-The theorem follows from the definitions of lsubst and lreplace; note that language substitutions over-approximate string substititons.
-\end{proof}
+The proof proceeds by structural induction on $r$.
+The base case ($r=\alpha, \alpha \in \Sigma$) is trivial.
+The choice and sequential cases require lemma 1, and prove a more general property that $\lsubst{r}{s_1}{s_2} \in \lang{r'}$ for some $r'$ isomorphic to $\lreplace{r}{r_1}{r_2}$.
+The closure case follows from the induction hypothesis applied to any $n$-unwinding.
 
@@
 %\end{description}
+\end{proof}
 
@@
 \end{proof}
+
 \begin{lem}
@@
 
-\begin{thm}{Safety and Sanitation Correctness for the String Fragment of P.} \label{thm:scorrect}
+\begin{thm}{Safety for the String Fragment of P.}
   Let $S$ be a term in the source language. If $\sctx \vdash S : \stringin{r}$ 
-  then $S \sreduces \rstr{s}$ and $\sctx \vdash \rstr{s} : \stringin{r}$; or else $S \err$.
+  then $S \sreduces \rstr{s}$ and $\rstr{s} : \stringin{r}$, or else $S \err$.
 \end{thm}
 \begin{proof}
-By induction on the typing relation, where (a) case holds by lemma \ref{thm:regexprops} in the S-T-Concat case and lemma \ref{thm:substcorrespondence} in the S-T-Replace case..
-The (b) cases hold by unstated, but standard, error propagation rules.
+By induction on the derivation of $\sctx \vdash S : \psi$. The interesting case
+is S-T-Replace, which requires lemma \ref{thm:substcorrespondence}. The 
+Coercion lemma additionally requires the second property of lemma \ref{thm:regexprops}.
+
+%\begin{description}
+%\item[S-T-Stringin-I:]
+%If $S = \rsconcat{S_1}{S_2} : \stringin{r}$ then
+%$S \sreduces S$ by S-E-RStr, and $\sctx \vdash S : \psi$ by assumption.
+%
+%\item[S-T-Concat:] 
+%Suppose $S = \rsconcat{S_1}{S_2} : \stringin{r_1 r_2}$.
+%By inversion, $\sctx \vdash S_1 : \stringin{r_1}$ and $\sctx \vdash S_1 : \stringin{r_2}$.
+%It follows by induction that either $S_1 \err$, $S_2 \err$, 
+%or $S_1 \sreduces \rstr{s_1}$ and $S_2 \sreduces \rstr{s_2}$ for some $s_1 \in \lang{r_1}$ and $s_2 \in \lang{r_2}$.
+%In the latter case $S \reduces \rstr{s_1s_2}$ by S-E-Concat and $\sctx \vdash \rstr{s_1s_2} : \tstr{r_1r_2}$ by \ref{thm:regexprops}.
+%In the former cases, $S \err$.
+%
+%\item[S-T-Replace:a]
+%Suppose $S = \rreplace{r}{S_1}{S_2}$ and $\sctx \vdash S : \stringin{r'}$.
+%By inversion $\sctx \vdash S_1 : \stringin{r_1}$ and $\sctx \vdash S_2 : \stringin{r_2}$ such that $\lsubst{r}{r_1}{r_2} = r'$.
+%By induction, $S_1 \err$, $S_2 \err$ 
+%or $S_1 \sreduces \rstr{s_1}$ and $S_2 \sreduces \rstr{s_2}$ such that 
+%In the latter case, we know $\lreplace{r}{s_1}{s_2} \in
+%\lang{\lsubst{r}{r_1}{r_2}}$ by lemma \ref{thm:substcorrespondence};
+%therefore by S-E-Replace, $S \sreduces \rstr{s}$ such that  
+%
+%$s \in \lang{\lsubst{r}{r_1}{r_2}} = \lang{r'}$.
+%So by S-T-String-I, $\rstr{s} : \stringin{r'}$.
+%In the former cases, $S \err$.
+%
+%\item[S-T-Coerce:]
+%Suppose $S = \coerce{r}{S_1}$ and $S : \stringin{r}$.
+%By inversion, $\sctx \vdash S_1 : \stringin{r'}$.
+%By induction, $S_1 \err$ or $S_1 \sreduces \rstr{s}$.
+%In the former case $S \err$ by propagation rules.
+%In the latter case we have by property 2 of \ref{thm:regexprops} that $s \in \lang{r}$ or else
+%$s \not \in \lang{r}$.
+%If $s \in \lang{r}$ then $\rstr{s} : \stringin{r}$.
+%If $s \not \in \lang{r}$ then $S \err$. 
+%\end{description}
 \end{proof}
-
-In addition to safety, we proof a correctness result for $\lambdas$ which ensures that well-typed terms of regular string type are in the language associated with their type.
-
-\begin{thm}{Correctness of Input Sanitation for $\lambdas$.}\label{thm:scorrect}
-  If  $\sctx \vdash S : \stringin{r}$ and $S \sreduces \rstr{s}$ then $s \in \lang{r}$.
-\end{thm}
-\begin{proof}
-  Follows directly from type safety, canonical forms for $\lambdas$.
-\end{proof}
-
-%\begin{thm}{Correctness of Input Sanitation (CIS) for $\lambdas$.}\label{thm:scorrect}
-%  If $\sctx \vdash S : \stringin{r}$ then either (a) $S \sreduces \rstr{s}$ for $s \in \lang{r}$
-%  or (b) $S \err$.
-%\end{thm}
 
@@
   Let $P$ be a term in the target language. If $\tctx \vdash P : \theta$ 
-  then $P \sreduces P'$ and $\tctx \vdash P' : \theta$, or else $P \err$.
+  then $P \sreduces P'$ and $\tctx \vdash P' : \theta$.
 \end{thm}
@@
 
-\begin{thm}{Translation Correctness.} \label{thm:tcorrect}
-If $\sctx \vdash S : \stringin{r}$ then there exists a $P$ such that $\trden{S} = P$ and either: 
-(a) $P \treduces \tstr{s}$ and $S \sreduces \strin{s}$, or
+We now present the main correctness result.
+
+\begin{thm}
+If $S : \strin{r}$ then there exists a P such that $\trden{s} = P$ and either: 
+(a) $P \treduces \tstr{s}$ and $S \sreduces \strin{s}$, and $s \in lang{r}$; or
 (b) $P \err$ and $S \err$.
@@
 choice of $P$; in each case, the choice is obvious. The subcases (a) proceed
-by inversion and appeals to our type safety theorems as well as the induction hypothesis. The subcases (b) proceed
+by inversion and appeals to the induction hypothesis. The subcases (b) proceed
 by the standard error propagation rules omitted for space.
@@
 are necessary.
-\end{proof}
-
-Finally, our main theorem establishes that input sanitation correctness of $\lambdas$ is preserved under the translation into $\lambdap$.
-
-\begin{thm}{Correctness of Input Sanitation for Translated Terms.}
-  If $\trden{S} = P$ and $\sctx \vdash S : \stringin{r}$ then either $P \err$ or $P \sreduces \tstr{s}$
-  for $s \in \lang{r}$.
-\end{thm}
-\begin{proof}
-  By theorem \ref{thm:tcorrect}, $P \sreduces \tstr{s}$ implies that $S \sreduces \strin{s}$.
-  By theorem \ref{thm:scorrect}, this together with the assumption that $S$ is well-typed implies that $s \in \lang{r}$.
-\end{proof}
-% PROOF OF TRANSLATION CORRECTNESS.
-% This proof is slightly out of say. It's SUFFICIENT, but is actually more
-% than sufficient because we prove \ref{scorrect} along with translation
-% correctness....
+
 %\begin{description}
@@
 
-\section{Implementation}
-
-An implementation of a similar system as a type system extension in the programming language Ace is discussed in \cite{fulton12} and \cite{fulton13}.
-The implementation only supports a special case of replacement where unsafe substrings are simply stripped.
-In practice, most libraries and frameworks escape command sequences instead of stripping characters; therefore, we plan to extend our implementation with
-support for the replace operation as described in this paper..
+\section{Implementation in {Atlang}}
+\begin{figure}
+\begin{lstlisting}
+@fn
+def sanitize(s : string_in[r'.*']):
+  return (s.replace(r'"', '&quot;') # TODO: is this right?
+           .replace(r'<', '&lt;')
+           .replace(r'>', '&gt;'))
+
+@fn
+def results_query(s : string_in[r'[^"]*']):
+  return 'SELECT * FROM users WHERE name="' + s + '"'
+
+@fn
+def results_div(s : string_in[r'[^<>]*']):
+  return '<div>Results for ' + s + '</div>'
+
+def main(db):
+  input = sanitize(user_input())
+  results = db.execute(results_query(input))
+  return results_div(input) + format(results)
+\end{lstlisting}
+\vspace{-10px}
+\caption{Regular string types in atlang, a library that enables static type checking for Python. }
+\end{figure}
+\begin{figure}
+\begin{lstlisting}
+class string_in(atlang.Type):
+  def __init__(self, rx):
+    rx = rx_normalize(rx)
+    atlang.Type.__init__(idx=rx)
+
+  def ana_Str(self, ctx, node):
+    if not in_lang(node.s, self.idx):
+      raise atlang.TypeError("...", node)
+
+  def trans_Str(self, ctx, node):
+    return astx.copy(node)
+
+  def syn_BinOp_Add(self, ctx, node):
+    left_t = ctx.syn(node.left)
+    right_t = ctx.syn(node.right)
+    if isinstance(left_t, string_in):
+      left_rx = left_t.idx
+      if isinstance(right_t, string_in):
+        right_rx = right_t.idx
+        return string_in[lconcat(left_rx, right_rx)]
+    raise atlang.TypeError("...", node)
+
+  def trans_BinOp_Add(self, ctx, node):
+    return astx.copy(node)
+
+  def syn_Method_replace(self, ctx, node):
+    [rx, exp] = node.args
+    if not isinstance(rx, ast.Str):
+      raise atlang.TypeError("...", node)
+    rx = rx.s
+    exp_t = ctx.syn(exp)
+    if not isinstance(exp_t, string_in):
+      raise atlang.TypeError("...", node)
+    exp_rx = exp_t.idx
+    return string_in[lreplace(self.idx, rx, exp_rx)]
+
+  def trans_Method_replace(self, ctx, node):
+    return astx.quote(
+      """__import__(re); re.sub(%0, %1, %2)""",
+      astx.Str(s=node.args[0]),
+      astx.copy(node.func.value),
+      astx.copy(node.args[1]))
+
+  def syn_Method_check(self, ctx, node):
+    [rx] = node.args
+    if not isinstance(rx, ast.Str):
+      raise atlang.TypeError("...", node)
+    return string_in[rx.s]
+
+  def trans_Method_check(self, ctx, node):
+    return astx.quote(
+      """__import__(string_in_helper);
+      string_in_helper.coerce(%0, %1)""",
+      astx.Str(s=other_t.idx),
+      astx.copy(node))
+
+  def check_Coerce(self, ctx, node, other_t):
+    # coercions can only be defined between 
+    # types with the same type constructor, 
+    if rx_sublang(other_t.idx, self.idx):
+      return other_t
+    else: raise atlang.TypeError("...", node)
+\end{lstlisting}
+\vspace{-10px}
+\caption{Implementation of the \texttt{string\_in} type constructor in atlang.}
+\end{figure}
+
+\begin{figure}
+\begin{lstlisting}
+output of successful compilation
+\end{lstlisting}
+\caption{Output of successful compilation.}
+\end{figure}
+
+\begin{figure}
+\begin{lstlisting}
+output of failed compilation
+\end{lstlisting}
+\caption{Output of failed compilation.}
+\end{figure}
+wwwww
+Here, we use the argument annotation syntax introduced in Python 3 (a similar syntax is available for Python 2.6+, not shown).
 
@@
 Composable analyses which complement existing approaches constitute a promising approach toward the integration of security concerns into programming languages.
-In this paper, we presented a system with both of these properties and defined a security-preserving transformation.
-Unlike other approaches, our solution complements existing, familiar solutions while providing a strong guarantee that traditional library and framework-based approaches are implemented and utilized correctly.
+In this paper, we presented a system with both of these properties and defined a security-prserving transformation.
+Unlike other approaches, our solution complements existing, familiar solutions while providing a strong guarantee that traditiona library and framework-based approaches are implemented and utilized correctly.
+
+Papers that needs to be cited in this section:
+\begin{itemize}
+  \item Ur/Web OSDI paper
+  \item Jif?
+  \item OWASP
+  \item XDuce and related papers. 
+  \item src?
+  \item Ace or Wyvern paper?
+  \item hotsos?
+  \item Haskell extension paper
+  \item Maybe some popular FOSS libraries/frameworks that do input sanitation?
+\end{itemize}
 
@@
 %
-\bibliographystyle{abbrv}
-
-\bibliography{research}
-
-% The bibliography should be embedded for final submission.
-
-%\begin{thebibliography}
-%
-%\bibitem{lamport94}
-%Leslie Lamport,
-%\emph{\LaTeX: a document preparation system}.
-%Addison Wesley, Massachusetts,
-%2nd edition,
-%1994.
-%
-%\end{thebibliography}
+%\bibliographystyle{abbrv}
+%
+%% The bibliography should be embedded for final submission.
+%
+%\bibliography{../research}
+%%\softraggedright
+%%P. Q. Smith, and X. Y. Jones. ...reference text...
+%
+%
+
 \end{document}

### 2014-08-07 sanitation-psp14/sanitation-psp14-short.tex

--- 2014-08-06 sanitation-psp14/sanitation-psp14-short.tex
@@
 %Evidence suggests that programmers are reluctant to adopt new languages to
-Web applications must ultimately generate strings that contain commands to be consumed by systems like web browsers and database engines.
-If these strings are constructed from user input that has not been properly sanitized, this can expose costly injection vulnerabilities.
+Web applications ultimately generate strings containing commands, which are   executed by systems like web browsers and database engines.
+Strings constructed from user input that has not been properly sanitized can thus cause command injection vulnerabilities.
 %Structured frameworks for constructing commands centralizes control over string sanitization, but verifying that the framework itself sanitizes user input correctly can still be tricky.
 
-In this paper, we introduce \emph{regular string types}, which classify strings known statically to be in a specified regular language.
-The language of a string is tracked by the type system through operations like concatenation, substitution and coercion, so regular string types can be used to implement, in essentially a conventional manner, the parts of a web application or application framework that constructs  commands.
-Simple type annotations at key points can be used to statically verify that sanitization has been performed correctly without introducing redundant run-time checks.
-
-We take the position that to be practical, such a type system cannot require that programmers adopt a new programming language.
-Ideally, such a ``special-purpose'' type system would be distributed as a library that could safely be used together with other such type systems.
-We support this by specifying a sound translation to a simple language containing only strings and regular expressions, then discuss implementing the type system together with this translation as a library in \texttt{atlang}, an extensible static type system for Python (being developed by the authors).% represents the most sophisticated application of this library to date.
+In this paper, we introduce \emph{regular string types}, which classify strings known statically to be in a specified regular language and support operations like concatenation, substitution and coercion.
+Regular string types can be used to implement, in essentially a conventional manner, the parts of a web application or application framework that construct such command strings.
+Straightforward type annotations at key interfaces can be used to statically verify that sanitization has been performed correctly without introducing redundant run-time checks.
+We specify this type system as a minimal typed lambda calculus, $\lambda_{RS}$.
+
+To be practical, adopting a type system like this should not require adopting a new programming language.
+Instead, we favor extensible type systems: new static type systems like this should be distributed as libraries atop a mechanism that guarantees that they can be safely composed.
+We support this by 1) specifying a  translation from $\lambda_{RS}$ to a language containing only strings and regular expressions, then, taking Python as such a language, 2) implement the type system together with the translation as a library using \texttt{atlang}, an extensible static type system for Python (being developed by the authors).% represents the most sophisticated application of this library to date.
 % Security-oriented programming languages do not enjoy wide adoption.
@@
 \section{Introduction}\label{intro}
-Improper input sanitation is a leading cause of security vulnerabilities in web applications \cite{OWASP}.
-Command injection attacks exploit improper input sanitation by inserting malicious code into an otherwise benign command.
-Modern web frameworks, libraries, and database abstraction layers attempt to ensure proper sanitation of user input.
-When these methods are unavailable or insufficient, developers implement custom sanitation techniques.
-In both cases, sanitation algorithms are implemented using the language's regular expression capabilities and usually \emph{replace} potentially unsafe strings with equivalent escaped strings.
-
-In this paper, we present a type system for implementing and statically checking input sanitation techniques.
+Command injection vulnerabilities are among the most common and severe in modern web applications \cite{OWASP}.
+They arise because web applications, at their boundaries, must control external systems that expose string-based command interfaces.
+For example, web browsers are controlled using HTML and Javascript sent from a server as a string, and database engines execute SQL queries also sent as strings.
+When these commands contain data derived from user input, care must be taken to ensure that the user cannot provide an input that will subvert the intended command.
+For example, a  SQL query constructed using string concatenation exposes a SQL injection vulnerability if \verb|name| is controlled by a user:
+\begin{lstlisting}[numbers=none]
+'SELECT * FROM users WHERE name="' + name + '"'
+\end{lstlisting}
+If a malicious user enters the name \lstinline{'"; DROP TABLE users --'}, the entire database could be erased.
+
+To avoid this problem, the program must \emph{sanitize} user input.
+For example, in this case, the developer (or, more often, a framework) might define a function \verb|sanitize| that prepends double quotes and existing backslashes  with a backslash, which SQL treats safely.
+Note that this function is not idempotent, so it should only be called once.
+Guaranteeing that user input has already been sanitized before it is used to construct a command is challenging.
+
+We observe that most such sanitization techniques can be understood in terms of \emph{regular languages}.
+For example, \verb|name| should be a string in the language described by the regular expression \verb!([^"\]|(\")|(\\))*!
+-- a sequence consisting of characters other than quotation marks or backslashes, or escaped quotation marks or escaped backslashes.
+Concrete syntax like this can be understood to desugar, in a standard way, to the abstract syntax for regular expressions shown in Figure \ref{fig:regex}.
+We will work with this ``core'' for simplicity, and assume basic familiarity with regular expressions \cite{cinderella}.
+
+In this paper, we present a static type system that tracks the regular language a string belongs to.
+We take advantage of closure and decidability properties of regular languages to support a number of useful operations on values of such \emph{regular string types}.
+These make it possible to implement sanitation protocols like the one just described in an essentially conventional manner.
+The result is a system where the fact that a string has been \emph{correctly} sanitized becomes manifest in its type.
+Missing calls to sanitization functions are detected statically, and, importantly, so are \emph{incorrectly implemented sanitization functions} (i.e.
+these functions need not be trusted).
+Run-time checks are only used when going from less precise to more precise types (e.g.
+at the edges of the system, where user input has not yet been validated) and no additional space overhead is required.
+%implementing and statically checking input sanitation techniques.
 Our solution suggests a more general approach to the integration of security concerns into programming language design.
@@
 
+TODO: need to update remainder of intro ...
 To demonstrate this approach, we present a simply typed lambda calculus with \emph{constrained strings}; that is, a set of string types parameterized by regular expressions.
@@
 \caption{Regular expressions over the alphabet $\Sigma$.}
+\label{fig:regex}
 \end{grammar}
@@
 
-\section{Implementation}
+\section{Implementation in {Atlang}}
+\begin{figure}
+\begin{lstlisting}
+@fn
+def sanitize(s : string_in[r'.*']):
+  return (s.replace(r'"', '&quot;') # TODO: is this right?
+           .replace(r'<', '&lt;')
+           .replace(r'>', '&gt;'))
+
+@fn
+def results_query(s : string_in[r'[^"]*']):
+  return 'SELECT * FROM users WHERE name="' + s + '"'
+
+@fn
+def results_div(s : string_in[r'[^<>]*']):
+  return '<div>Results for ' + s + '</div>'
+
+def main(db):
+  input = sanitize(user_input())
+  results = db.execute(results_query(input))
+  return results_div(input) + format(results)
+\end{lstlisting}
+\vspace{-10px}
+\caption{Regular string types in atlang, a library that enables static type checking for Python. }
+\end{figure}
+\begin{figure}
+\begin{lstlisting}
+class string_in(atlang.Type):
+  def __init__(self, rx):
+    rx = rx_normalize(rx)
+    atlang.Type.__init__(idx=rx)
+
+  def ana_Str(self, ctx, node):
+    if not in_lang(node.s, self.idx):
+      raise atlang.TypeError("...", node)
+
+  def trans_Str(self, ctx, node):
+    return astx.copy(node)
+
+  def syn_BinOp_Add(self, ctx, node):
+    left_t = ctx.syn(node.left)
+    right_t = ctx.syn(node.right)
+    if isinstance(left_t, string_in):
+      left_rx = left_t.idx
+      if isinstance(right_t, string_in):
+        right_rx = right_t.idx
+        return string_in[lconcat(left_rx, right_rx)]
+    raise atlang.TypeError("...", node)
+
+  def trans_BinOp_Add(self, ctx, node):
+    return astx.copy(node)
+
+  def syn_Method_replace(self, ctx, node):
+    [rx, exp] = node.args
+    if not isinstance(rx, ast.Str):
+      raise atlang.TypeError("...", node)
+    rx = rx.s
+    exp_t = ctx.syn(exp)
+    if not isinstance(exp_t, string_in):
+      raise atlang.TypeError("...", node)
+    exp_rx = exp_t.idx
+    return string_in[lreplace(self.idx, rx, exp_rx)]
+
+  def trans_Method_replace(self, ctx, node):
+    return astx.quote(
+      """__import__(re); re.sub(%0, %1, %2)""",
+      astx.Str(s=node.args[0]),
+      astx.copy(node.func.value),
+      astx.copy(node.args[1]))
+
+  def syn_Method_check(self, ctx, node):
+    [rx] = node.args
+    if not isinstance(rx, ast.Str):
+      raise atlang.TypeError("...", node)
+    return string_in[rx.s]
+
+  def trans_Method_check(self, ctx, node):
+    return astx.quote(
+      """__import__(string_in_helper);
+      string_in_helper.coerce(%0, %1)""",
+      astx.Str(s=other_t.idx),
+      astx.copy(node))
+
+  def check_Coerce(self, ctx, node, other_t):
+    # coercions can only be defined between 
+    # types with the same type constructor, 
+    if rx_sublang(other_t.idx, self.idx):
+      return other_t
+    else: raise atlang.TypeError("...", node)
+\end{lstlisting}
+\vspace{-10px}
+\caption{Implementation of the \texttt{string\_in} type constructor in atlang.}
+\end{figure}
+
+\begin{figure}
+\begin{lstlisting}
+output of successful compilation
+\end{lstlisting}
+\caption{Output of successful compilation.}
+\end{figure}
+
+\begin{figure}
+\begin{lstlisting}
+output of failed compilation
+\end{lstlisting}
+\caption{Output of failed compilation.}
+\end{figure}
+wwwww
+Here, we use the argument annotation syntax introduced in Python 3 (a similar syntax is available for Python 2.6+, not shown).
+
+\section{Conclusion}
 

### 2014-08-09 sanitation-psp14/sanitation-psp14-short.tex
Changed description of sanitation.
No longer motivates the fact that sanitation can be difficult, but this
paragraph should be in-line with our actual definition of sanitation found
later in the paper.

@@
 To avoid this problem, the program must \emph{sanitize} user input. 
-For example, in this case, the developer (or, more often, a framework) might define a function \verb|sanitize| that prepends double quotes and existing backslashes  with a backslash, which SQL treats safely.
+For example, in this case, the developer (or, more often, a framework) might define a function \verb|sanitize| that escapes any double quotes, which are necessary in order to force uesr input to be interpreted as anything other than a string literal.
+%prepends double quotes and existing backslashes  with a backslash, which SQL treats safely.
 Note that this function is not idempotent, so it should only be called once.
@@
 \begin{lstlisting}
+#todo: why are these r's here?
+#TODO-nrf this is maybe kind-of right bue I think there's a nicer way to do it.
 @fn
 def sanitize(s : string_in[r'.*']):
-  return (s.replace(r'"', '&quot;') # TODO: is this right?
+  return (s.replace(r'"', '&quot;') 
            .replace(r'<', '&lt;')

### 2014-08-11 sanitation-psp14/sanitation-psp14-short.tex
Small changes.

@@
 another string; for instance, $\lsubst{a|b}{a}{c} = c$.
-In order to computer the type resulting from aubstitution, we also need to compute
-the result of replacing on language with another inside a given language.
+In order to compute the type resulting from aubstitution, we also need to compute
+the result of replacing one language with another inside a given language.
 Finally, just for convienance, we provide a coerce operation. The introduction of
@@
 % is modified before and after this figure.
+
+\ 
+\newpage
+\ 
+\newpage
+\
 
@@
 
+The main theorem of this paper is Theorem \ref{thm:main}.
+Establishing this result requires an additional theorem (Theorem \ref{tcorrect}), which establishes a relationship between canonical forms for the string fragments of $\lambdas$ and $\lambdap$.
+
 \begin{thm}{Translation Correctness.} \label{thm:tcorrect}
@@
 
-\begin{thm}{Correctness of Input Sanitation for Translated Terms.}
+\begin{thm}{Correctness of Input Sanitation for Translated Terms.}\label{thm:main}
   If $\trden{S} = P$ and $\sctx \vdash S : \stringin{r}$ then either $P \err$ or $P \sreduces \tstr{s}$
@@
 
-\section{Conclusion}
-
-An implementation of a similar system as a type system extension in the programming language Ace is discussed in \cite{fulton12} and \cite{fulton13}.
-The implementation only supports a special case of replacement where unsafe substrings are simply stripped.
+In this section, we present an example atlang program using our extension together
+with our implementation of the extension in Atlang modulo ${\sf lreplace}$.
+Note that in addition to Figure \ref{fig:regex}, regular expressions in atlang
+also support the {\sf \^} character, which is complementation. Complementation
+is definable in terms of Figure \ref{fig:regex} whenever $\Sigma$ is
+fixed.
+
+TODO-nrf insert a pargraph or two which explains what's going on...
+
+\subsection{An Example}
+
+Figure \ref{fig:example} demosntrates the use of string types in atlang.
+The sanitize function takes an arbitrary string and returns a string without
+double quotes and left and right brackets. In this example, we use HTML
+escape sequences; many variants of SQL provide their own syntax
+for this as well.
+
+TODO-nrf line by line of example.
+
+\subsection{Extension Implementation}
+
+TODO-nrf line by line of extension implementation as ewll.
+
+An implementation of a similar extension as the one presented here is discussed in \cite{fulton12} and \cite{fulton13}.
+Our prior work only supports a special case of replacement where unsafe substrings are completely removed.
 In practice, most libraries and frameworks escape command sequences instead of stripping characters; therefore, we plan to extend our implementation with
-support for the replace operation as described in this paper..
+support for the replace operation as described in this paper.
 

### 2014-08-13 sanitation-psp14/sanitation-psp14-short.tex
finished intro.

@@
 To avoid this problem, the program must \emph{sanitize} user input. 
-For example, in this case, the developer (or, more often, a framework) might define a function \verb|sanitize| that escapes any double quotes, which are necessary in order to force uesr input to be interpreted as anything other than a string literal.
-%prepends double quotes and existing backslashes  with a backslash, which SQL treats safely.
+For example, in this case, the developer (or, more often, a framework) might define a function \verb|sanitize| that 
+%escapes any double quotes, which are necessary in order to force uesr input to be interpreted as anything other than a string literal.
+prepends double quotes and existing backslashes  with a backslash, which SQL treats safely.
 Note that this function is not idempotent, so it should only be called once.
@@
 
-TODO: need to update remainder of intro ...
-To demonstrate this approach, we present a simply typed lambda calculus with \emph{constrained strings}; that is, a set of string types parameterized by regular expressions.
-If $s : \stringin{r}$, then $s$ is a string matching the language $r$.
-Additionally, we include an operation $\rreplace{r}{s_1}{s_2}$ which corresponds to the replace mechanism available in most regular expression libraries; that is, any substring of $s_1$ matching $r$ is replaced with $s_2$.
-The type of this expression is the computed, and is likely ``smaller" or more constrained than the type of $s_1$.
-Libraries, frameworks or functions which construct and execute commands containing input can specify a safe subset $\stringin{r_{\sf spec}}$ of strings, and input sanitation algorithms can construct such a string using ${\sf rreplace}$ or, optionally, by coercion (in which case a runtime check is inserted).
-We also show how this system is translated into a host language containing a regular expression library such that the safety guarantee of the extended language is preserved.
-
-Summarily, we present a simple type system extension which ensures the absence of input sanitation vulnerabilities by statically checking input sanitation algorithms which use an underlying regular expression library.
-This approach is \emph{composable} in the sense that it is a conservative extension.
-This approach is also \emph{complementary} to existing input sanitation techniques which use string replacement for input sanitation.
+Our type system for regular string types, $\lambdas$, is defined as an exntenstion to a simply-typed lambda calculus.
+Opeartions on strings are given types according to corresponding, type-level
+computations on regular expressions. A concatenation operation is defined in 
+order to demosntrate this principle, but the premier operation of our system
+is \emph{replacement}, which captures a rich set of input sanitation techniques
+including escaped characters and encodings. In addition to defining this system,
+we present a correctness-preserving compilation of $\lambdas$ into an underlying
+language with ordinary strings as a regular expression library.
+
+In addition to this theoretical contribution, we also explain how our extension
+may be implemented in the Ace Programming Language, which is itself an extension
+of Python -- a popular web programming language.
+
+Summarily, we present a simple type system extension which ensures the absence
+of input sanitation vulnerabilities by statically checking the correctness and 
+correct use of input sanitation algorithms. This approach is \emph{composable}
+in the sense that it is a conservative extension. This approach is \emph{complementary}
+to exsiting input sanitation techniques that use regular expressions for input santiation.
 
@@
 
+The Wyvern programming language provides a general framework for composing lanugage extensions \cite{tsl}\cite{wyvern}.
+Our work identifies one particular extension, and is therefore complementary to Wyvern and related
+work on extensible programming languages.
 We are also unaware of any extensible programming languages which emphasize applications to security concerns.
@@
 Note that in addition to Figure \ref{fig:regex}, regular expressions in atlang
-also support the {\sf \^} character, which is complementation. Complementation
-is definable in terms of Figure \ref{fig:regex} whenever $\Sigma$ is
-fixed.
+also support complementation using the carot.
 

### 2014-08-13 sanitation-psp14/sanitation-psp14-short.tex
Added changes.

@@
 % Names of type systems presented in paper
-\newcommand{\lcs}{\lambda_{CS}}
+\newcommand{\lcs}{\lambda_{S}}
 
@@
 Straightforward type annotations at key interfaces can be used to statically verify that sanitization has been performed correctly without introducing redundant run-time checks.
-We specify this type system as a minimal typed lambda calculus, $\lambda_{RS}$.
+We specify this type system as a minimal typed lambda calculus, $\lambda_{S}$.
 
@@
 Instead, we favor extensible type systems: new static type systems like this should be distributed as libraries atop a mechanism that guarantees that they can be safely composed.
-We support this by 1) specifying a  translation from $\lambda_{RS}$ to a language containing only strings and regular expressions, then, taking Python as such a language, 2) implement the type system together with the translation as a library using \texttt{atlang}, an extensible static type system for Python (being developed by the authors).% represents the most sophisticated application of this library to date.
+We support this by 1) specifying a  translation from $\lambda_{S}$ to a language containing only strings and regular expressions, then, taking Python as such a language, 2) implement the type system together with the translation as a library using \texttt{atlang}, an extensible static type system for Python (being developed by the authors).% represents the most sophisticated application of this library to date.
 % Security-oriented programming languages do not enjoy wide adoption.
@@
 \begin{itemize}
-  \item Page 3 contains a definition of $\lambdas, \lambdap$ and the translation from $\lambdas$ to $\lambdap$.
+  \item Page 3 contains a definition of $\lambdas, \lambdap$ and the translation from $\lambdas$ to $\lambdap$. Grammar follows immediately at the top of page 4.
   \item In \S 2.1 we state some properties about regular expressions which are needed in our correctness proofs.
@@
 %%support this assertion.
+
+%The $\lcs$ language gives static semantics for common regular expression library
+%functions. In this treatment, we include concatenation and filtering.
+%The ${\tt filter}$ function removes all instances of a regular expression in a string,
+%while concatenation ($+$) concatenates two strings.
+%
+%\subsection{Typing}
+%
+%The $\strin[r]$ type is parameterized by regular expressions; if $e:\strin[r]$,
+%then $e \in r$. Mapping from an arbitrary $\str$ to a $\strin[r]$ requires
+%defining an algorithm -- in terms of filter -- for converting a $\strin[.*]$
+%into a $\strin[r]$. The static semantics of the language defines the types of
+%operations on regular expressions in terms of well-understood properties about
+%regular lanugages; we recall these properties in section 3.
+%
+%\subsection{Dynamics}
+%
+%There are two evaluation judgements: $e:T \reduces e'$ and $e:T \treduces i$.
+%The $\reduces$ relation is between $\lcs$ expressions, while the $\treduces$
+%relation is a mapping from $\lcs$ expressions into internal language expressions
+%$i$ such that $i \ival$.
+%
+%Safety of the evaluation relation depends upon an injective mapping from $\lcs$ types info
+%internal language types. This relation, $h$, is defined below.
+%
+%\subsection{Type Safety}
+%
+%The type safety proof relies upon some assumptions about the type system and
+%dynamics of the internal language, as well as some properties of regular
+%languages.
+%
+%There must exist a translation from $\lcs$ types to the types of the internal
+%language. For the remainder of this paper, we call the type translation function
+%$h$.
+%
+%\begin{defn}[Type Translation Function $h$]
+%  The type translation function $h : Type \rightarrow IType$ is defined as follows:
+%  \begin{itemize}
+%    \item $\forall r. h(\strin[r]) = istr$
+%    \item $h(\str) = istr$
+%  \end{itemize}
+%\end{defn}
+%
+%Additionally, we assume that the internal language contains an implementation of strings,
+%together with operations for concatenation and filtering by regular expression.
+%
+%\begin{defn}[Types of internal values]
+%  Let $\tstrf{s}$ range over string literals and $r$ over regular expressions.
+%  Internal values are typed as follows:
+%\begin{itemize}
+%  \item If $e=\tstrf{s}$ then $e:istr$.
+%  \item If $e=\ifilter{r}{\tstrf{s}}$ then $e:istr$.
+%  \item If $e=\tstrf{s_1}+\tstrf{s_2}$ then $e:istr$.
+%\end{itemize}
+%\end{defn}
+%
+%For simplicity, we assume a fixed translation from
+%$\lcs$ regular expressions to regular expressions recognizable by the internal
+%language's regex library (in practice, a fixed translation is acceptable.)
+%To summarize, we assume an internal language containing 
+%a string type together with operations for string concatentation and filtering. We
+%expect closure over strings for both operations.
+%Finally, recall that ${\tt ifilter}$ is only needed for dynamic casts, which may be removed without
+%descreasing the expressivity or even usability of the language.
+%Finally, the semantics of the filter function are defined in terms of ${\tt rl\_filter}$,
+%which is a static version of ${\tt ifilter}$.
+%
+
+% I can't achieve the correct latout here. We should have something like:
+%       Lam_S Synax         Lam_P Syntax
+%       Lam_S Sematics      LAm_P Semantics
+%                Translation rules
+% Or, alternatively, we can flip reflect things so that the entire page
+% reads top-down (Lam_S syntax Lam_S smantics horizontally aligned w/ one
+% another).
+
+% Unfortunately, I can't get that to work. So, we're settling for something
+% slightly ugly, where the fitures appear in the incorrect order and the 
+% translation rules are on top.
+
+% The translations rules *just won't* go to the bottom of the page, so 
+% instead I am placing them at the top. Since the figures are out of order,
+% it's even more important that we get the number gith. therefore, the counter
+% is modified before and after this figure.
+
+\addtocounter{figure}{4}
+\begin{figure*}[t]\label{fig:trans}
+
+$\fbox{\inferrule{}{\trden{S} = P}}$
+\begin{mathpar}
+
+\inferrule[Tr-string]
+{ \ }
+{\trden{\strin{s}} = \tstr{s}}
+
+\inferrule[Tr-Concat]
+{ \trden{S_1} = P_1 \\ \trden{S_2} = P_2}
+{\trden{\rsconcat{S_1}{S_2}} = \tconcat{P_1}{P_2}}
+
+\inferrule[Tr-Subst]
+{ \trden{S_1} = P_1 \\ \trden{S_2} = P_2 }
+{\trden{ \sisubst{r}{S_1}{S_2} } = \tsubst{\rx{r}}{P_1}{P_2} }
+
+\inferrule[Tr-Coerce-Ok]
+{ S : \strin{r} \\ \lang{r'} \subseteq \lang{r}}
+{\trden{ \coerce{r'}{S} } = \tstr{s}}
+
+\inferrule[Tr-Coerce-NotOk]
+{\trden{S} = P \\ S : \strin{r} \\ \lang{r'} \not \subseteq \lang{r}}
+{\trden{\coerce{r'}{S}} = \tcheck{\rx{r'}}{P}}
+
+\end{mathpar}
+\caption{Translation from source terms (S) to target terms (P).
+The translation is type-directed in the Tr-Coerce cases.}
+\end{figure*}
+\addtocounter{figure}{-5}
+
+%
+% Begin typing/translation relation
+%
+
+\newcommand{\sctx}{\Psi} % Context for external typing
+\newcommand{\tctx}{\Theta} % Context for internal typing
+
+\newcommand{\ereduces}{\Downarrow}
+
+\begin{figure}[t]\label{fig:lambdas}
+\small
+$\fbox{\inferrule{}{\sctx \vdash S : \psi}}$
+~~~~$\sctx ::= \emptyset \pipe \sctx, x : \psi$
+\begin{mathpar}
+%\inferrule[S-T-Abs]
+%{\sctx, x : T_1 \vdash e : T_2}
+%{\sctx \vdash \lambda x.e : T_1 \rightarrow T_2}
+%
+%\inferrule[S-T-App]
+%{\sctx \vdash e_1 : T' \\ \sctx \vdash e_2 : T' \rightarrow T}
+%{\sctx \vdash e_1 e_2 : T}
+%  
+\inferrule[S-T-Stringin-I]
+{s \in \lang{r}}
+{\sctx \vdash \strin{s} : \stringin{r}}
+
+\inferrule[S-T-Concat]
+{\sctx \vdash S_1 : \stringin{r_1} \\ \sctx \vdash S_2 : \stringin{r_2}}
+{\sctx \vdash \rsconcat{S_1}{S_2} : \stringin{r_1 \cdot r_2}}
+
+\inferrule[S-T-Replace]
+{\sctx \vdash S_1 : \stringin{r_1} \\ \sctx \vdash S_2 : \stringin{r_2} \\\\ \lreplace{r}{r_1}{r_2} = r'}
+{\sctx \vdash \sisubst{r}{S_1}{S_2} : \stringin{r'}}
+
+\inferrule[S-T-coerce]
+{\sctx \vdash S : \stringin{r'}}
+{\sctx \vdash \coerce{r}{S} : \stringin{r}}
+
+\end{mathpar}
+
+\caption{Typing rules for our fragment of $\lambdas$. 
+The typing context $\sctx$ is standard.}
+\label{fig:etyping}
+\end{figure}
+
+\begin{figure}[t]
+\small
+$\fbox{\inferrule{}{S \sreduces S}}$
+$\fbox{\inferrule{}{S \err}}$
+\begin{mathpar}
+%
+%\inferrule[S-E-Abs]                                                             
+%{ \ }                                                                           
+%{\lambda x.e \sreduces \lambda x.e}                                             
+%
+%\inferrule[S-E-App]
+%{ e_1 \sreduces e_1' \\ e_2 \sreduces e_2' }
+%{ e_1 e_2 \sreduces e_1' e_2'}
+%
+\inferrule[S-E-RStr]
+{ \ }
+{\strin{s} \sreduces \strin{s}}
+
+\inferrule[S-E-Concat]
+{S_1 \sreduces \strin{s_1} \\ S_2 \sreduces \strin{s_2}} 
+{\rsconcat{S_1}{S_2} \sreduces \strin{s_1 s_2}} % ???
+
+\inferrule[S-E-Replace]
+{S_1 \sreduces \strin{s_1} \\ S_2 \sreduces \strin{s_2} \\ \lsubst{r}{s_1}{s_2} = s} 
+{\sisubst{r}{S_1}{S_2} \sreduces \sistr{s}}
+
+\inferrule[S-E-Coerce-OK]
+{S \sreduces \strin{s} \\ s \in \lang{r}}
+{\coerce{r}{S} \sreduces \strin{s}}
+
+\inferrule[S-E-Coerce-Err]
+{S \sreduces \strin{s} \\ s \not \in \lang{r}}
+{\coerce{r}{S} \err}
+
+\end{mathpar}
+\caption{Big step semantics for our fragment of $\lambdas$.
+Error propagation rules are omitted.}
+\label{fig:eeval}
+\end{figure}
+
+\newpage
+
+\begin{figure}[t]\label{fig:lambdap}
+\small
+$\fbox{\inferrule{}{\tctx \vdash P : \theta}}$
+~~~~$\tctx ::= \emptyset \pipe \tctx, x : \theta$
+
+\begin{mathpar}
+%\inferrule[P-T-Abs]
+%{\tctx, x : T_1 \vdash e : T_2}
+%{\tctx \vdash \lambda x.e : T_1 \rightarrow T_2}
+%
+%\inferrule[P-T-App]
+%{\tctx \vdash e_1 : T \\ \tctx \vdash e_2 : T' \rightarrow T}
+%{\tctx \vdash e_1 e_2 : T}
+%
+\inferrule[P-T-String]
+{ \ }
+{\tctx \vdash \tstr{s} : \str}
+
+\inferrule[P-T-Regex]
+{ \ }
+{\tctx \vdash \rx{r} : \regex}
+
+\inferrule[P-T-Concat]
+{\tctx \vdash P_1 : \str \\ \tctx \vdash P_2 : \str}
+{\tctx \vdash \tconcat{P_1}{P_2} : \str}
+
+\inferrule[P-T-Replace]
+{\tctx \vdash P_1 : \regex \\ \tctx \vdash P_2 : \str \\ \tctx \vdash P_3 : \str }
+{\tctx \vdash \preplace{P_1}{P_2}{P_3} : \str}
+
+\inferrule[P-T-Check]
+{\tctx \vdash P_1 : \regex \\ \tctx \vdash P_2 : \str \\ }
+{\tctx \vdash \tcheck{P_1}{P_2} : \str}
+\end{mathpar}
+\caption{Typing rules for our fragment of $\lambdap$.
+The typing context $\tctx$ is standard.}
+\label{fig:etyping}
+\end{figure}
+
+\begin{figure}[t]
+\small
+$\fbox{\inferrule{}{P \treduces P}}$
+$\fbox{\inferrule{}{P \err}}$
+
+\begin{mathpar}
+
+%\inferrule[P-E-Abs]
+%{ \ }
+%{\lambda x.e \sreduces \lambda x.e}
+%
+%\inferrule[P-E-App]
+%{ e_1 \sreduces e_1' \\  e_2 \sreduces e_2' }
+%{ e_1 e_2 \sreduces e_1' e_2'}
+%
+  \inferrule[P-E-Str]
+{ \ }
+{\tstr{s} \treduces \tstr{s}}
+
+\inferrule[P-E-Rx]
+{ \ }
+{\rx{r} \treduces \rx{r}}
+
+\inferrule[P-E-Concat]
+{P_1 \treduces \tstr{s_1} \\ P_2 \treduces \tstr{s_2}} 
+{\tconcat{P_1}{P_2} \treduces \tstr{s_1 s_2}} % ???
+
+\inferrule[P-E-Replace]
+{P_1 \treduces \rx{r} \\ P_2 \treduces \tstr{s_2} \\ P_3 \treduces \tstr{s_3} \\ \lsubst{r}{s_2}{s_3} = s} 
+{\preplace{P_1}{P_2}{P_3} \treduces \tstr{s}}
+
+\inferrule[P-E-Check-OK]
+{P_1 \treduces \rx{r} \\ P_2 \treduces \strin{s} \\ s \in \lang{r}}
+{\tcheck{P_1}{P_2} \treduces \tstr{s}}
+
+\inferrule[P-E-Check-Err]
+{P_1 \treduces \rx{r} \\ P_2 \treduces \tstr{s} \\ s \not \in \lang{r}}
+{\tcheck{P_1}{P_2} \err}
+\end{mathpar}
+\caption{Big step semantics for our fragment of $\lambdap$.
+Error propagation rules are omitted.}
+\label{fig:ieval}
+\end{figure}
+
+\newpage
+\ 
+\newpage
+\ 
+\newpage
+\
 
@@
 
-%The $\lcs$ language gives static semantics for common regular expression library
-%functions. In this treatment, we include concatenation and filtering.
-%The ${\tt filter}$ function removes all instances of a regular expression in a string,
-%while concatenation ($+$) concatenates two strings.
-%
-%\subsection{Typing}
-%
-%The $\strin[r]$ type is parameterized by regular expressions; if $e:\strin[r]$,
-%then $e \in r$. Mapping from an arbitrary $\str$ to a $\strin[r]$ requires
-%defining an algorithm -- in terms of filter -- for converting a $\strin[.*]$
-%into a $\strin[r]$. The static semantics of the language defines the types of
-%operations on regular expressions in terms of well-understood properties about
-%regular lanugages; we recall these properties in section 3.
-%
-%\subsection{Dynamics}
-%
-%There are two evaluation judgements: $e:T \reduces e'$ and $e:T \treduces i$.
-%The $\reduces$ relation is between $\lcs$ expressions, while the $\treduces$
-%relation is a mapping from $\lcs$ expressions into internal language expressions
-%$i$ such that $i \ival$.
-%
-%Safety of the evaluation relation depends upon an injective mapping from $\lcs$ types info
-%internal language types. This relation, $h$, is defined below.
-%
-%\subsection{Type Safety}
-%
-%The type safety proof relies upon some assumptions about the type system and
-%dynamics of the internal language, as well as some properties of regular
-%languages.
-%
-%There must exist a translation from $\lcs$ types to the types of the internal
-%language. For the remainder of this paper, we call the type translation function
-%$h$.
-%
-%\begin{defn}[Type Translation Function $h$]
-%  The type translation function $h : Type \rightarrow IType$ is defined as follows:
-%  \begin{itemize}
-%    \item $\forall r. h(\strin[r]) = istr$
-%    \item $h(\str) = istr$
-%  \end{itemize}
-%\end{defn}
-%
-%Additionally, we assume that the internal language contains an implementation of strings,
-%together with operations for concatenation and filtering by regular expression.
-%
-%\begin{defn}[Types of internal values]
-%  Let $\tstrf{s}$ range over string literals and $r$ over regular expressions.
-%  Internal values are typed as follows:
-%\begin{itemize}
-%  \item If $e=\tstrf{s}$ then $e:istr$.
-%  \item If $e=\ifilter{r}{\tstrf{s}}$ then $e:istr$.
-%  \item If $e=\tstrf{s_1}+\tstrf{s_2}$ then $e:istr$.
-%\end{itemize}
-%\end{defn}
-%
-%For simplicity, we assume a fixed translation from
-%$\lcs$ regular expressions to regular expressions recognizable by the internal
-%language's regex library (in practice, a fixed translation is acceptable.)
-%To summarize, we assume an internal language containing 
-%a string type together with operations for string concatentation and filtering. We
-%expect closure over strings for both operations.
-%Finally, recall that ${\tt ifilter}$ is only needed for dynamic casts, which may be removed without
-%descreasing the expressivity or even usability of the language.
-%Finally, the semantics of the filter function are defined in terms of ${\tt rl\_filter}$,
-%which is a static version of ${\tt ifilter}$.
-%
-
-% I can't achieve the correct latout here. We should have something like:
-%       Lam_S Synax         Lam_P Syntax
-%       Lam_S Sematics      LAm_P Semantics
-%                Translation rules
-% Or, alternatively, we can flip reflect things so that the entire page
-% reads top-down (Lam_S syntax Lam_S smantics horizontally aligned w/ one
-% another).
-
-% Unfortunately, I can't get that to work. So, we're settling for something
-% slightly ugly, where the fitures appear in the incorrect order and the 
-% translation rules are on top.
-
-% The translations rules *just won't* go to the bottom of the page, so 
-% instead I am placing them at the top. Since the figures are out of order,
-% it's even more important that we get the number gith. therefore, the counter
-% is modified before and after this figure.
-
-\ 
-\newpage
-\ 
-\newpage
-\
-
-\addtocounter{figure}{4}
-\begin{figure*}\label{fig:trans}
-
-$\fbox{\inferrule{}{\trden{S} = P}}$
-\begin{mathpar}
-
-\inferrule[Tr-string]
-{ \ }
-{\trden{\strin{s}} = \tstr{s}}
-
-\inferrule[Tr-Concat]
-{ \trden{S_1} = P_1 \\ \trden{S_2} = P_2}
-{\trden{\rsconcat{S_1}{S_2}} = \tconcat{P_1}{P_2}}
-
-\inferrule[Tr-Subst]
-{ \trden{S_1} = P_1 \\ \trden{S_2} = P_2 }
-{\trden{ \sisubst{r}{S_1}{S_2} } = \tsubst{\rx{r}}{P_1}{P_2} }
-
-\inferrule[Tr-Coerce-Ok]
-{ S : \strin{r} \\ \lang{r'} \subseteq \lang{r}}
-{\trden{ \coerce{r'}{S} } = \tstr{s}}
-
-\inferrule[Tr-Coerce-NotOk]
-{\trden{S} = P \\ S : \strin{r} \\ \lang{r'} \not \subseteq \lang{r}}
-{\trden{\coerce{r'}{S}} = \tcheck{\rx{r'}}{P}}
-
-\end{mathpar}
-\caption{Translation from source terms (S) to target terms (P).
-The translation is type-directed in the Tr-Coerce cases.}
-\end{figure*}
-\addtocounter{figure}{-5}
-
-%
-% Begin typing/translation relation
-%
-
-\newcommand{\sctx}{\Psi} % Context for external typing
-\newcommand{\tctx}{\Theta} % Context for internal typing
-
-\newcommand{\ereduces}{\Downarrow}
-
-\begin{figure}[t]\label{fig:lambdas}
-\small
-$\fbox{\inferrule{}{\sctx \vdash S : \psi}}$
-~~~~$\sctx ::= \emptyset \pipe \sctx, x : \psi$
-\begin{mathpar}
-%\inferrule[S-T-Abs]
-%{\sctx, x : T_1 \vdash e : T_2}
-%{\sctx \vdash \lambda x.e : T_1 \rightarrow T_2}
-%
-%\inferrule[S-T-App]
-%{\sctx \vdash e_1 : T' \\ \sctx \vdash e_2 : T' \rightarrow T}
-%{\sctx \vdash e_1 e_2 : T}
-%  
-\inferrule[S-T-Stringin-I]
-{s \in \lang{r}}
-{\sctx \vdash \strin{s} : \stringin{r}}
-
-\inferrule[S-T-Concat]
-{\sctx \vdash S_1 : \stringin{r_1} \\ \sctx \vdash S_2 : \stringin{r_2}}
-{\sctx \vdash \rsconcat{S_1}{S_2} : \stringin{r_1 \cdot r_2}}
-
-\inferrule[S-T-Replace]
-{\sctx \vdash S_1 : \stringin{r_1} \\ \sctx \vdash S_2 : \stringin{r_2} \\\\ \lreplace{r}{r_1}{r_2} = r'}
-{\sctx \vdash \sisubst{r}{S_1}{S_2} : \stringin{r'}}
-
-\inferrule[S-T-coerce]
-{\sctx \vdash S : \stringin{r'}}
-{\sctx \vdash \coerce{r}{S} : \stringin{r}}
-
-\end{mathpar}
-
-\caption{Typing rules for our fragment of $\lambdas$. 
-The typing context $\sctx$ is standard.}
-\label{fig:etyping}
-\end{figure}
-
-\begin{figure}[t]
-\small
-$\fbox{\inferrule{}{S \sreduces S}}$
-$\fbox{\inferrule{}{S \err}}$
-\begin{mathpar}
-%
-%\inferrule[S-E-Abs]                                                             
-%{ \ }                                                                           
-%{\lambda x.e \sreduces \lambda x.e}                                             
-%
-%\inferrule[S-E-App]
-%{ e_1 \sreduces e_1' \\ e_2 \sreduces e_2' }
-%{ e_1 e_2 \sreduces e_1' e_2'}
-%
-\inferrule[S-E-RStr]
-{ \ }
-{\strin{s} \sreduces \strin{s}}
-
-\inferrule[S-E-Concat]
-{S_1 \sreduces \strin{s_1} \\ S_2 \sreduces \strin{s_2}} 
-{\rsconcat{S_1}{S_2} \sreduces \strin{s_1 s_2}} % ???
-
-\inferrule[S-E-Replace]
-{S_1 \sreduces \strin{s_1} \\ S_2 \sreduces \strin{s_2} \\ \lsubst{r}{s_1}{s_2} = s} 
-{\sisubst{r}{S_1}{S_2} \sreduces \sistr{s}}
-
-\inferrule[S-E-Coerce-OK]
-{S \sreduces \strin{s} \\ s \in \lang{r}}
-{\coerce{r}{S} \sreduces \strin{s}}
-
-\inferrule[S-E-Coerce-Err]
-{S \sreduces \strin{s} \\ s \not \in \lang{r}}
-{\coerce{r}{S} \err}
-
-\end{mathpar}
-\caption{Big step semantics for our fragment of $\lambdas$.
-Error propagation rules are omitted.}
-\label{fig:eeval}
-\end{figure}
-
-\newpage
-
-\begin{figure}[t]\label{fig:lambdap}
-\small
-$\fbox{\inferrule{}{\tctx \vdash P : \theta}}$
-~~~~$\tctx ::= \emptyset \pipe \tctx, x : \theta$
-
-\begin{mathpar}
-%\inferrule[P-T-Abs]
-%{\tctx, x : T_1 \vdash e : T_2}
-%{\tctx \vdash \lambda x.e : T_1 \rightarrow T_2}
-%
-%\inferrule[P-T-App]
-%{\tctx \vdash e_1 : T \\ \tctx \vdash e_2 : T' \rightarrow T}
-%{\tctx \vdash e_1 e_2 : T}
-%
-\inferrule[P-T-String]
-{ \ }
-{\tctx \vdash \tstr{s} : \str}
-
-\inferrule[P-T-Regex]
-{ \ }
-{\tctx \vdash \rx{r} : \regex}
-
-\inferrule[P-T-Concat]
-{\tctx \vdash P_1 : \str \\ \tctx \vdash P_2 : \str}
-{\tctx \vdash \tconcat{P_1}{P_2} : \str}
-
-\inferrule[P-T-Replace]
-{\tctx \vdash P_1 : \regex \\ \tctx \vdash P_2 : \str \\ \tctx \vdash P_3 : \str }
-{\tctx \vdash \preplace{P_1}{P_2}{P_3} : \str}
-
-\inferrule[P-T-Check]
-{\tctx \vdash P_1 : \regex \\ \tctx \vdash P_2 : \str \\ }
-{\tctx \vdash \tcheck{P_1}{P_2} : \str}
-\end{mathpar}
-\caption{Typing rules for our fragment of $\lambdap$.
-The typing context $\tctx$ is standard.}
-\label{fig:etyping}
-\end{figure}
-
-\begin{figure}[t]
-\small
-$\fbox{\inferrule{}{P \treduces P}}$
-$\fbox{\inferrule{}{P \err}}$
-
-\begin{mathpar}
-
-%\inferrule[P-E-Abs]
-%{ \ }
-%{\lambda x.e \sreduces \lambda x.e}
-%
-%\inferrule[P-E-App]
-%{ e_1 \sreduces e_1' \\  e_2 \sreduces e_2' }
-%{ e_1 e_2 \sreduces e_1' e_2'}
-%
-  \inferrule[P-E-Str]
-{ \ }
-{\tstr{s} \treduces \tstr{s}}
-
-\inferrule[P-E-Rx]
-{ \ }
-{\rx{r} \treduces \rx{r}}
-
-\inferrule[P-E-Concat]
-{P_1 \treduces \tstr{s_1} \\ P_2 \treduces \tstr{s_2}} 
-{\tconcat{P_1}{P_2} \treduces \tstr{s_1 s_2}} % ???
-
-\inferrule[P-E-Replace]
-{P_1 \treduces \rx{r} \\ P_2 \treduces \tstr{s_2} \\ P_3 \treduces \tstr{s_3} \\ \lsubst{r}{s_2}{s_3} = s} 
-{\preplace{P_1}{P_2}{P_3} \treduces \tstr{s}}
-
-\inferrule[P-E-Check-OK]
-{P_1 \treduces \rx{r} \\ P_2 \treduces \strin{s} \\ s \in \lang{r}}
-{\tcheck{P_1}{P_2} \treduces \tstr{s}}
-
-\inferrule[P-E-Check-Err]
-{P_1 \treduces \rx{r} \\ P_2 \treduces \tstr{s} \\ s \not \in \lang{r}}
-{\tcheck{P_1}{P_2} \err}
-\end{mathpar}
-\caption{Big step semantics for our fragment of $\lambdap$.
-Error propagation rules are omitted.}
-\label{fig:ieval}
-\end{figure}
-
-\newpage
-\ 
-\newpage
-\ 
-\newpage
+\subsection{Definition of $\lambdas$}
+
+The $\lambdas$ system extends the simply-typed lambda calculus with regular
+expression types; an explanation of significant typing rule follows:
+\begin{itemize}
+  \item Rule S-T-Concat states that the type of concatenated strings is 
+    obtained by concatenating the regular expressions for each string.
+  \item Rule S-T-Replace defines the type of $\textsf{rreplace}$. The expression $\textsf{rreplace}$
+    evaluates to the result of substituting every string matching $r$ in $s_1$
+    with $s_2$. This operation corresponds to the \textsf{str_replace} function of PHP.
+    The type of these expressions is defined in terms of an extra-linguistic \textsf{lreplace}
+    function, which is defined later in this section.
+    The expression $\lreplace{r}{r_1}{r_2}$ is obtained by
+    starting with $r_1$, and replacing any subexpression matching $r$ with $r_2$.
+  \item Rule S-T-Coerce allows coercion; however, note that this coercion cannot
+    invalidate our safety property because we ensure appropraite checks are always
+    inserted wherever coercions are used.
+\end{itemize}
+
+\subsection{Definition of $\lambdap$}
+
+The system $\lambdap$ is a stright-forward extension of a simply typed lambda claculus
+with a string type and a regular epxression tpye. We include two operations which are available in the regular expression
+library of any modern programming language. The check operation ensures that an expression
+recognizes a string, and the replace operation is string replacement.
+
+\subsection{Definition of Translation}
+
+The translation from $\lambdas$ to $\lambdap$ is defined in figure 5.
+The coercion cases are most interesting. If the safety of coercion in manifest in the
+types of the expressions, then no runtime check is inserted.
+If the safety of coercion is not manifest in the types, then a check is inserted.
+
+In practice, the type of a replacement rarely matches a specification.
+Therefore, it is convienant in an implementation to always insert the appropriate
+coercion, and then only raise type errors when an automatically inserted coercion
+actually requires the insertion of a runtime check. Alternatively, this policy may 
+be codified in the type system itself using subtyping \cite{fulton12}.
 
@@
 
+In this section, we establish type safety for the source ($\lambdas$) and
+target ($\lambdap$) languages. In addition to type safety, we also prove
+a stronger correctness property for $\lambdas$.
+
+Our first two theorems establish that our rules preserve the well-formedness of
+regular expressions. The standard lemmas required for safety of the simply
+typed lambda claculus are also required; proofs for these do not differ substntially
+from \cite{pfpl}.
+
 \begin{lem}
@@
 We now prove safety for the string fragment of the source and target languages.
-
-\begin{thm}{Safety and Sanitation Correctness for the String Fragment of P.} \label{thm:scorrect}
+The easiest property is type safety of $\lambdap$, which follows almost directly from
+type safety for the simply typed lambda calculus.
+
+\begin{thm} 
+  Let $P$ be a term in the target language. If $\tctx \vdash P : \theta$ 
+  then $P \sreduces P'$ and $\tctx \vdash P' : \theta$, or else $P \err$.
+\end{thm}
+
+Safety for the string fragment is more involved, because it involves validating that
+the type system's definition is justified by our theorems about regular languages.
+Note that type safety does \emph{not} guarantee that strings of a type are in the
+correct language. We isolate this property so that we may reason about its preservation
+under translation to $\lambdap$, which does not have regular expression types.
+
+\begin{thm}{Type Safety for the String Fragment of $\lambdas$.} \label{thm:scorrect}
   Let $S$ be a term in the source language. If $\sctx \vdash S : \stringin{r}$ 
@@
 
-In addition to safety, we proof a correctness result for $\lambdas$ which ensures that well-typed terms of regular string type are in the language associated with their type.
+In addition to safety, $\lambdas$ requires a correctness result ensuring that well-typed terms of regular string type are in the language associated with their type.
 
@@
 
-%\begin{thm}{Correctness of Input Sanitation (CIS) for $\lambdas$.}\label{thm:scorrect}
-%  If $\sctx \vdash S : \stringin{r}$ then either (a) $S \sreduces \rstr{s}$ for $s \in \lang{r}$
-%  or (b) $S \err$.
-%\end{thm}
-
-\begin{thm} 
-  Let $P$ be a term in the target language. If $\tctx \vdash P : \theta$ 
-  then $P \sreduces P'$ and $\tctx \vdash P' : \theta$, or else $P \err$.
-\end{thm}
-
 \subsection{Translation Correctness}
 
-The main theorem of this paper is Theorem \ref{thm:main}.
-Establishing this result requires an additional theorem (Theorem \ref{tcorrect}), which establishes a relationship between canonical forms for the string fragments of $\lambdas$ and $\lambdap$.
+The main theorem of this paper is Theorem \ref{thm:main},
+which establishes that Theorem \ref{thm:scorrect} is preserved under translation
+into the target language $\lambdap$.
+
+Establishing this result requires an additional theorem establishing a relationship between canonical forms for the string fragments of $\lambdas$ and $\lambdap$.
 
@@
   By theorem \ref{thm:tcorrect}, $P \sreduces \tstr{s}$ implies that $S \sreduces \strin{s}$.
-  By theorem \ref{thm:scorrect}, this together with the assumption that $S$ is well-typed implies that $s \in \lang{r}$.
+  By theorem \ref{thm:scorrect}, the above property together with the assumption that $S$ is well-typed implies that $s \in \lang{r}$.
 \end{proof}
@@
 \section{Implementation in {Atlang}}
-\begin{figure}
+\begin{figure}\label{fig:example}
 \begin{lstlisting}
-#todo: why are these r's here?
-#TODO-nrf this is maybe kind-of right bue I think there's a nicer way to do it.
 @fn
@@
 \caption{Regular string types in atlang, a library that enables static type checking for Python. }
-\end{figure}
+\end{figure}\label{fig:impl}
 \begin{figure}
@@
 \end{figure}
-wwwww
-Here, we use the argument annotation syntax introduced in Python 3 (a similar syntax is available for Python 2.6+, not shown).
 
 In this section, we present an example atlang program using our extension together
-with our implementation of the extension in Atlang modulo ${\sf lreplace}$.
+with our implementation of the extension in Atlang (modulo ${\sf lreplace}$).
 Note that in addition to Figure \ref{fig:regex}, regular expressions in atlang
-also support complementation using the carot.
-
-TODO-nrf insert a pargraph or two which explains what's going on...
-
-\subsection{An Example}
-
-Figure \ref{fig:example} demosntrates the use of string types in atlang.
-The sanitize function takes an arbitrary string and returns a string without
-double quotes and left and right brackets. In this example, we use HTML
-escape sequences; many variants of SQL provide their own syntax
-for this as well.
-
-TODO-nrf line by line of example.
-
-\subsection{Extension Implementation}
-
-TODO-nrf line by line of extension implementation as ewll.
+also support complementation using the \^{} symbol. Atlang is implemented in Python,
+and shares the syntax of Python (including the argument annotation syntax introduced in Python 3);
+however, Atlang is statically typed and supported the definition of type system extensions by 
+overriding Python's dynamic dispatch protocol.
 
@@
 support for the replace operation as described in this paper.
+
+\subsection{An Example}
+
+Figure \ref{fig:example} demosntrates the use of string types in atlang.
+The sanitize function takes an arbitrary string and returns a string without
+double quotes or left and right brackets. In this example, we use HTML
+escape sequences.
+
+The main function receives user input and passes this input to a 
+sanitize function, which replaces all double quotes and brackets with HTML
+escape sequences.
+
+The result of applying sanitize to input is apped to two functions which
+construct a safe query and safe output.
+The arguments to the result and output construction functions constitute \emph{specifications}.
+In the case of \textsf{results_query}, this specification ensures that user input is always interpreted
+as a string literal by the SQL server.
+In the case of \textsf{results_div}, this specification ensures that user input does 
+not contain any HTML tags, which is a conservative but effective policy for preventing
+XSS attacks.
+
+Note that ${\sf input}$ does not actually meet these specifications without additional machinery.
+The type of ${\sf input}$ is quite large
+and does not actually equal the specified domains of the query or output construction
+methods. This mismatch is common -- in fact, nearly universal. Therefore, 
+our implementation includes a simple subtyping relation between regular expression
+types. 
+This subtyping relation is justified theoretically by the fact that language inclusion is
+decidable; see \cite{fulton12} for a formal definition of the subtyping relation.
+Additionally, our extension remains composable because subtyping is defined on a type-by-type basis;
+see \cite{fulton13} for a discussion of subtyping in Atlang (referred to there as Ace).
+
+\subsection{Implementation of the Regular Expression Type}
+
+We implemented a variation on the type system presented in this paper. The only
+significant difference is that we only support replacements where $s_2$ is the empty
+string. Therefore, our implementation respects the system presented in section 2
+only modulo the definition of \textsf{lreplace}.
+
+Atlang translates programs using type definitions, which may extend both the
+static and dynamic semantics of the language. New types are defined as Python
+classes; figure \ref{fig:impl} contains the source code of our implementation.
+
+The \textsf{string_in} type has an indexing regular expression \texttt{idx}.
+Our translation is defined by the \textsf{trans_} methods while the \textsf{syn_}
+methods define our type checker. The atlang type checker or translater defer
+to these methods whenever an expression of type \textsf{string_in} is encountered.
 

### 2014-08-13 sanitation-psp14/sanitation-psp14.tex
Small changes.
Added changes.

@@
 
+\subsection{Definitiosn and Translation}
+
 \subsection{Safety of the Source and Target Languages}
+
+In this section, we establish type safety for the source ($\lambdas$) and
+target ($\lambdap$) languages. In addition to type safety, we also prove
+a stronger correctness property for $\lambdas$.
+
+Our first two theorems establish that our rules preserve the well-formedness of
+regular expressions. The standard lemmas required for safety of the simply
+typed lambda claculus are also required; proofs for these do not differ substntially
+from \cite{pfpl}.
 
@@
 
-We now prove safety for the string fragment of the source and target languages.
-
-\begin{thm}{Safety for the String Fragment of P.}
+Type safety for $\lambdas$ í
+
+\begin{thm}{Safety for the String Fragment of $\lambdas$.}
   Let $S$ be a term in the source language. If $\sctx \vdash S : \stringin{r}$ 
@@
 
-\bibliography{../research}
+\begin{thebibliography}
+
+    @article{greenwade93,
+          author  = "George D. Greenwade",
+              title   = "The {C}omprehensive {T}ex {A}rchive {N}etwork ({CTAN})",
+                  year    = "1993",
+                      journal = "TUGBoat",
+                          volume  = "14",
+                              number  = "3",
+                                  pages   = "342--351"
+                                }
+
+\end{thebibliography}
+
 %\softraggedright

### 2014-08-13 sanitation-psp14/sanitation-psp14-short.tex
typos

@@
 
-Our type system for regular string types, $\lambdas$, is defined as an exntenstion to a simply-typed lambda calculus.
-Opeartions on strings are given types according to corresponding, type-level
+Our type system for regular string types, $\lambdas$, is defined as an exntension to a simply-typed lambda calculus.
+Operations on strings are given types according to corresponding, type-level
 computations on regular expressions. A concatenation operation is defined in 
-order to demosntrate this principle, but the premier operation of our system
+order to demonstrate this principle, but the premier operation of our system
 is \emph{replacement}, which captures a rich set of input sanitation techniques
@@
 in the sense that it is a conservative extension. This approach is \emph{complementary}
-to exsiting input sanitation techniques that use regular expressions for input santiation.
+to existing input sanitation techniques that use regular expressions for input sanitation.
 
@@
 In this section, we explain how our approach to the input sanitation problem differs from each of these approaches.
-More important than these differences, however, is our more general assertion that language extensibility is a promising approach toward consideration of security goals in programming lanugage design.
+More important than these differences, however, is our more general assertion that language extensibility is a promising approach toward consideration of security goals in programming language design.
 
@@
 
-The Wyvern programming language provides a general framework for composing lanugage extensions \cite{tsl}\cite{wyvern}.
+The Wyvern programming language provides a general framework for composing language extensions \cite{tsl}\cite{wyvern}.
 Our work identifies one particular extension, and is therefore complementary to Wyvern and related
@@
 Incorporating regular expressions into the type system is not novel.
-The XDuce system \cite{HosoyaVouillonPierce2000ICFP,HosoyaPierce2002} checks XML documents against schemas using regular expressions.
+The XDuce system \cite{HosoyaVouillonPierce2000ICFP,HosoyaPierce2002} checks XML documents against schema using regular expressions.
 Similarly, XHaskell \cite{xhaskell} focuses on XML documents.
@@
   \item Our system is defined within an extensible type system.
-  \item We demonstrate that regular expression types are applicable to the web security domain, whereas previous work on regular expression types focused on XML schemas.
+  \item We demonstrate that regular expression types are applicable to the web security domain, whereas previous work on regular expression types focused on XML schema.
   \item Although our static replacement operation is definable in some languages with regular expression types, we are the first to expose this operation and connect the semantics of regular language replacement with the semantics of string substitution via a type safety and compilation correctness argument.
@@
 another string; for instance, $\lsubst{a|b}{a}{c} = c$.
-In order to compute the type resulting from aubstitution, we also need to compute
+In order to compute the type resulting from substitution, we also need to compute
 the result of replacing one language with another inside a given language.
-Finally, just for convienance, we provide a coerce operation. The introduction of
+Finally, just for convenience, we provide a coerce operation. The introduction of
 coercion requires handling of runtime errors.
@@
    \item In \S 2.2 we prove type safety for $\lambdap$ as well as both type safety and correctness for $\lambdas$.
-  \item In \S 2.3 we prove that translation preserves the correctness reesult about $\lambdas$.
+  \item In \S 2.3 we prove that translation preserves the correctness result about $\lambdas$.
 \end{itemize}
@@
 
-The system $\lambdap$ is a stright-forward extension of a simply typed lambda claculus
-with a string type and a regular epxression tpye. We include two operations which are available in the regular expression
+The system $\lambdap$ is a straight-forward extension of a simply typed lambda calculus
+with a string type and a regular expression tpye. We include two operations which are available in the regular expression
 library of any modern programming language. The check operation ensures that an expression
@@
 \begin{defn}[$\tt{lsubst}$]
-The replation $\lsubst{r}{s_1}{s_2} = s$ produces a string $s$ in which all substrings of $s_1$ matching $r$ are replaced with $s_2$.
+The relation $\lsubst{r}{s_1}{s_2} = s$ produces a string $s$ in which all substrings of $s_1$ matching $r$ are replaced with $s_2$.
 \end{defn}
@@
 \begin{proof}
-The theorem follows from the definitions of lsubst and lreplace; note that language substitutions over-approximate string substititons.
+The theorem follows from the definitions of lsubst and lreplace; note that language substitutions over-approximate string substitutions.
 \end{proof}
@@
 
-The result of applying sanitize to input is apped to two functions which
+The result of applying sanitize to input is appended to two functions which
 construct a safe query and safe output.
@@
 Our translation is defined by the \textsf{trans_} methods while the \textsf{syn_}
-methods define our type checker. The atlang type checker or translater defer
+methods define our type checker. Atlang defers type checking and translation
 to these methods whenever an expression of type \textsf{string_in} is encountered.

### 2014-08-13 sanitation-psp14/sanitation-psp14-short.tex
updated

--- 2014-08-13 sanitation-psp14/sanitation-psp14-short.tex
@@
 %Evidence suggests that programmers are reluctant to adopt new languages to
-Web applications ultimately generate strings containing commands, which are   executed by systems like web browsers and database engines.
-Strings constructed from user input that has not been properly sanitized can thus cause command injection vulnerabilities.
+Web applications must ultimately command systems like web browsers and database engines using strings.
+Strings constructed using user input that has not been properly sanitized can thus cause command injection vulnerabilities.
 %Structured frameworks for constructing commands centralizes control over string sanitization, but verifying that the framework itself sanitizes user input correctly can still be tricky.
-
-In this paper, we introduce \emph{regular string types}, which classify strings known statically to be in a specified regular language and support operations like concatenation, substitution and coercion.
-Regular string types can be used to implement, in essentially a conventional manner, the parts of a web application or application framework that construct such command strings.
-Straightforward type annotations at key interfaces can be used to statically verify that sanitization has been performed correctly without introducing redundant run-time checks.
+In this paper, we introduce \emph{regular string types}, which classify strings known statically to be in a specified regular language.
+These types come equipped with common operations like concatenation, substitution and coercion, so they  can be used to implement, in essentially a conventional manner, the parts of a web application or application framework that construct command strings.
+Simple type annotations at key interfaces can be used to statically verify that sanitization has been performed correctly without introducing redundant run-time checks.
 We specify this type system as a minimal typed lambda calculus, $\lambda_{RS}$.
 
-To be practical, adopting a type system like this should not require adopting a new programming language.
-Instead, we favor extensible type systems: new static type systems like this should be distributed as libraries atop a mechanism that guarantees that they can be safely composed.
+To be practical, adopting a specialized type system like this should not require the adoption of a new programming language.
+Instead, we favor extensible type systems: new type system fragments like this should be distributed as libraries atop a mechanism that guarantees that they can be safely composed.
 We support this by 1) specifying a  translation from $\lambda_{RS}$ to a language containing only strings and regular expressions, then, taking Python as such a language, 2) implement the type system together with the translation as a library using \texttt{atlang}, an extensible static type system for Python (being developed by the authors).% represents the most sophisticated application of this library to date.
@@
 \section{Introduction}\label{intro}
-Command injection vulnerabilities are among the most common and severe in modern web applications \cite{OWASP}.
-They arise because web applications, at their boundaries, must control external systems that expose string-based command interfaces.
+Command injection vulnerabilities are among the most common and severe security vulnerabilities in modern web applications \cite{OWASP}.
+They arise because web applications, at their boundaries, control external systems by issuing  commands represented using  strings.
 For example, web browsers are controlled using HTML and Javascript sent from a server as a string, and database engines execute SQL queries also sent as strings.
-When these commands contain data derived from user input, care must be taken to ensure that the user cannot provide an input that will subvert the intended command.
-For example, a  SQL query constructed using string concatenation exposes a SQL injection vulnerability if \verb|name| is controlled by a user:
+When these commands must include data derived from user input, care must be taken to ensure that the user cannot construct input that subverts the intended command.
+For example, a  SQL query constructed using string concatenation exposes a SQL injection vulnerability:
 \begin{lstlisting}[numbers=none]
@@
 To avoid this problem, the program must \emph{sanitize} user input.
-For example, in this case, the developer (or, more often, a framework) might define a function \verb|sanitize| that 
-%escapes any double quotes, which are necessary in order to force uesr input to be interpreted as anything other than a string literal.
-prepends double quotes and existing backslashes  with a backslash, which SQL treats safely.
+For example, in this case, the developer (or, more often, a framework) might define a function \verb|sanitize| that prepends double quotes (and existing backslashes)  with a backslash, which SQL treats as safely.
+Guaranteeing that user input has already been sanitized before it is used to construct a command is challenging.
 Note that this function is not idempotent, so it should only be called once.
-Guaranteeing that user input has already been sanitized before it is used to construct a command is challenging.
-
-We observe that most such sanitization techniques can be understood in terms of \emph{regular languages}.
+
+We observe that most such sanitization techniques can be understood in terms of \emph{regular languages} \cite{cinderella}.
 For example, \verb|name| should be a string in the language described by the regular expression \verb!([^"\]|(\")|(\\))*!
--- a sequence consisting of characters other than quotation marks or backslashes, or escaped quotation marks or escaped backslashes.
-Concrete syntax like this can be understood to desugar, in a standard way, to the abstract syntax for regular expressions shown in Figure \ref{fig:regex}.
-We will work with this ``core'' for simplicity, and assume basic familiarity with regular expressions \cite{cinderella}.
+-- a sequence of characters other than quotation marks and backslashes; these can only appear escaped.
+This concrete pattern syntax can be understood to desugar, in a standard way, to the syntax for regular expressions shown in Figure \ref{fig:regex}, where $r \cdot r$ is sequencing and $r + r$ is  disjunction.
+We will work with this ``core'' for simplicity.
 
 In this paper, we present a static type system that tracks the regular language a string belongs to.
+For example, the output of \verb|sanitize| will be in the regular language described by the regular expression above.
+We write this as $\mathcal{L}\{r\}$.
 We take advantage of closure and decidability properties of regular languages to support a number of useful operations on values of such \emph{regular string types}.
 These make it possible to implement sanitation protocols like the one just described in an essentially conventional manner.
-The result is a system where the fact that a string has been \emph{correctly} sanitized becomes manifest in its type.
+The result is a system where the fact that a string has been {correctly} sanitized is manifest in its type.
 Missing calls to sanitization functions are detected statically, and, importantly, so are \emph{incorrectly implemented sanitization functions} (i.e.
 these functions need not be trusted).
-Run-time checks are only used when going from less precise to more precise types (e.g.
-at the edges of the system, where user input has not yet been validated) and no additional space overhead is required.
+These guarantees require run-time checks only when going from less precise to more precise types (e.g.
+at the edges of the system, where user input has not yet been validated).
 %implementing and statically checking input sanitation techniques.
@@
 
-Our type system for regular string types, $\lambdas$, is defined as an exntenstion to a simply-typed lambda calculus.
-Opeartions on strings are given types according to corresponding, type-level
-computations on regular expressions. A concatenation operation is defined in 
-order to demosntrate this principle, but the premier operation of our system
-is \emph{replacement}, which captures a rich set of input sanitation techniques
-including escaped characters and encodings. In addition to defining this system,
-we present a correctness-preserving compilation of $\lambdas$ into an underlying
-language with ordinary strings as a regular expression library.
-
-In addition to this theoretical contribution, we also explain how our extension
-may be implemented in the Ace Programming Language, which is itself an extension
-of Python -- a popular web programming language.
-
-Summarily, we present a simple type system extension which ensures the absence
-of input sanitation vulnerabilities by statically checking the correctness and 
-correct use of input sanitation algorithms. This approach is \emph{composable}
-in the sense that it is a conservative extension. This approach is \emph{complementary}
-to exsiting input sanitation techniques that use regular expressions for input santiation.
+We will begin in Sec.
+\ref{calculus} by specifying this type system minimally, as a conservative extension of the simply typed lambda calculus called $\lambda_{RC}$.
+This allows us to specify the guarantees that the type system provides precisely.
+We also formally specify a translation from this calculus to a typed calculus with only standard strings and regular expressions, intending it as a guide to language implementors interested in building this feature into their own languages.
+This also demonstrates that no additional space overhead is required.
+
+Waiting for a language designer to build this feature in is unsatisfying.
+Moreover, we would also face a ``chicken-and-egg problem'': justifying its inclusion into a commonly used language requires empirically demonstrating that it is useful, but this is difficult to do if developers have no way to use it in practice.
+As such, we take the position that a better path forward for the community is to work within a programming language where such type system fragments can be introduced modularly and orthogonally, as libraries.
+
+In Sec.
+\ref{atlang}, we show how to implement the type system fragment we have specified  using \texttt{atlang}, an extensible static type system implemented as a library inside Python.
+\texttt{atlang} leverages local type inference to control the semantics of literal forms, so reguar string types can be introduced using string literals without any run-time overhead.
+Coercions that are known to be  safe due to a sublanguage relationship are performed implicitly, also without run-time overhead.
+This results in a \emph{usably secure} system: working with regular strings differs little from working with standard strings.
+
+We conclude after discussing related work in Sec. \ref{related}.
+
+% TODO: need to update remainder of intro ...
+To demonstrate this approach, we present a simply typed lambda calculus with \emph{constrained strings}; that is, a set of string types parameterized by regular expressions.
+If $s : \stringin{r}$, then $s$ is a string matching the language $r$.
+Additionally, we include an operation $\rreplace{r}{s_1}{s_2}$ which corresponds to the replace mechanism available in most regular expression libraries; that is, any substring of $s_1$ matching $r$ is replaced with $s_2$.
+The type of this expression is the computed, and is likely ``smaller" or more constrained than the type of $s_1$.
+Libraries, frameworks or functions which construct and execute commands containing input can specify a safe subset $\stringin{r_{\sf spec}}$ of strings, and input sanitation algorithms can construct such a string using ${\sf rreplace}$ or, optionally, by coercion (in which case a runtime check is inserted).
+We also show how this system is translated into a host language containing a regular expression library such that the safety guarantee of the extended language is preserved.
+
+% Summarily, we present a simple type system extension which ensures the absence of input sanitation vulnerabilities by statically checking input sanitation algorithms which use an underlying regular expression library.
+This approach is \emph{composable} in the sense that it is a conservative extension.
+This approach is also \emph{complementary} to existing input sanitation techniques which use string replacement for input sanitation.
 
@@
 
-\subsection{Related Work and Alternative Approaches}
-
-The input sanitation problem is well-understood.
-There exist a large number of techniques and technologies, proposed by both practitioners and researchers, for preventing injection-style attacks.
-In this section, we explain how our approach to the input sanitation problem differs from each of these approaches.
-More important than these differences, however, is our more general assertion that language extensibility is a promising approach toward consideration of security goals in programming lanugage design.
-
-Unlike \emph{frameworks and libraries} provided by languages such as Haskell and Ruby, our type system provides a \emph{static} guarantee that input is always properly sanitized before use.
-Doing so requires reasoning about the operations on regular languages corresponding to standard operations on strings; we are unaware of any production system which contains this form of reasoning.
-Therefore, even where frameworks and libraries provide a viable interface or wrapper around input sanitation, our approach is complementary because it ensures the correctness of the framework or library itself.
-Furthermore, our approach is more general than database abstraction layers because our mechanism is applicable to all forms of command injection (e.g.
-shell injection or remote file inclusion).
-
-A number of research languages provide static guarantees that a program is free of input sanitation vulnerabilities \cite{UrFlowOSDI10}. 
-Unlike this work, our solution to the input sanitation problem has a very low barrier to adoption; for instance, our implementation conservatively extends Python -- a popular language among web developers.
-We also believe our general approach is better-positioned for security, where continuously evolving threats might require frequent addition of new analyses; in these cases, the composability and generality of our approach is a substantial advantage.
-
-%\begin{itemize}
-%  \item Our system is a light-weight solution to a single class of sanitation vulnerabilities (e.g. we do not address Cross-Site Scripting).
-%  \item Our system is defined as a library in terms of an extensible type system, as opposed to a stand-alone language.
-Instead of introducing new technologies and methodologies for addressing security problems, we provide a light-weight static analysis which complements approaches developers already understand well.
-%  \item Our implementation of the translation is implemented in Python and shares its grammar.
-Since Python is a popular programming language among web developers, the barrier between our research and adopted technologies is lower than for greenfield security-oriented languages.
-%\end{itemize}
-
-The Wyvern programming language provides a general framework for composing lanugage extensions \cite{tsl}\cite{wyvern}.
-Our work identifies one particular extension, and is therefore complementary to Wyvern and related
-work on extensible programming languages.
-We are also unaware of any extensible programming languages which emphasize applications to security concerns.
-
-Incorporating regular expressions into the type system is not novel.
-The XDuce system \cite{HosoyaVouillonPierce2000ICFP,HosoyaPierce2002} checks XML documents against schemas using regular expressions.
-Similarly, XHaskell \cite{xhaskell} focuses on XML documents.
-We differ from this and related work in at least three ways:
-\begin{itemize}
-  \item Our system is defined within an extensible type system.
-  \item We demonstrate that regular expression types are applicable to the web security domain, whereas previous work on regular expression types focused on XML schemas.
-  \item Although our static replacement operation is definable in some languages with regular expression types, we are the first to expose this operation and connect the semantics of regular language replacement with the semantics of string substitution via a type safety and compilation correctness argument.
-\end{itemize}
-
-In conclusion, our contribution is a type system, implemented within an extensible type system, for checking the correctness of input sanitation algorithms.
-
 %\subsection{Outline}
@@
 
-\section{A Type System for String Sanitation}
-
-In this section we define a language for statically checked string sanitation ($\lambdas$).
+\section{Regular String Types, Minimally}\label{calculus}
+\renewcommand{\grammarlabel}[2]{#1\hfill#2}
+
+\begin{figure}[t]
+\begin{grammar}
+<$r$> ::= $\epsilon$ | $.$ | $a$ | $r \cdot r$ | $r + r$ | $r*$ \hfill $a \in \Sigma$
+
+\caption{Regular expressions over the alphabet $\Sigma$.}
+\label{fig:regex}
+\end{grammar}
+\end{figure}
+
+\begin{figure}[t]
+\begin{grammar}
+
+<$\sigma$> ::=  $\sigma \rightarrow \sigma$     \hfill  source types          \alt
+$\stringin{r}$
+
+<$e$> ::= 
+      $x$ \hfill source terms \alt 
+      $\lambda x {:} \sigma . e$ \alt
+      $e~(e)$ \alt
+      $\strin{s}$ \hfill $s \in \Sigma^{*}$ \alt
+      $\rsconcat{e}{e}$ \alt
+      $\sisubst{r}{e}{e}$ \alt
+      $\coerce{r}{e}$
+\caption{Syntax of $\lambda_{RS}$.}\label{fig:lamrs}
+\end{grammar}
+\end{figure}
+
+\newcommand{\sctx}{\Psi} % Context for external typing
+\newcommand{\tctx}{\Theta} % Context for internal typing
+
+\newcommand{\ereduces}{\Downarrow}
+
+\begin{figure}[t]\label{fig:lambdas}
+\small
+$\fbox{\inferrule{}{\sctx \vdash S : \sigma}}$
+~~~~$\sctx ::= \emptyset \pipe \sctx, x : \sigma$
+\begin{mathpar}
+%\inferrule[S-T-Abs]
+%{\sctx, x : T_1 \vdash e : T_2}
+%{\sctx \vdash \lambda x.e : T_1 \rightarrow T_2}
+%
+%\inferrule[S-T-App]
+%{\sctx \vdash e_1 : T' \\ \sctx \vdash e_2 : T' \rightarrow T}
+%{\sctx \vdash e_1 e_2 : T}
+%  
+\inferrule[S-T-Stringin-I]
+{s \in \lang{r}}
+{\sctx \vdash \strin{s} : \stringin{r}}
+
+\inferrule[S-T-Concat]
+{\sctx \vdash S_1 : \stringin{r_1} \\ \sctx \vdash S_2 : \stringin{r_2}}
+{\sctx \vdash \rsconcat{S_1}{S_2} : \stringin{r_1 \cdot r_2}}
+
+\inferrule[S-T-Replace]
+{\sctx \vdash S_1 : \stringin{r_1} \\ \sctx \vdash S_2 : \stringin{r_2} \\\\ \lreplace{r}{r_1}{r_2} = r'}
+{\sctx \vdash \sisubst{r}{S_1}{S_2} : \stringin{r'}}
+
+\inferrule[S-T-coerce]
+{\sctx \vdash S : \stringin{r'}}
+{\sctx \vdash \coerce{r}{S} : \stringin{r}}
+
+\end{mathpar}
+
+\caption{Typing rules for our fragment of $\lambdas$. 
+The typing context $\sctx$ is standard.}
+\label{fig:etyping}
+\end{figure}
+
+\begin{figure}[t]
+\small
+$\fbox{\inferrule{}{S \sreduces S}}$
+$\fbox{\inferrule{}{S \err}}$
+\begin{mathpar}
+%
+%\inferrule[S-E-Abs]                                                             
+%{ \ }                                                                           
+%{\lambda x.e \sreduces \lambda x.e}                                             
+%
+%\inferrule[S-E-App]
+%{ e_1 \sreduces e_1' \\ e_2 \sreduces e_2' }
+%{ e_1 e_2 \sreduces e_1' e_2'}
+%
+\inferrule[S-E-RStr]
+{ \ }
+{\strin{s} \sreduces \strin{s}}
+
+\inferrule[S-E-Concat]
+{S_1 \sreduces \strin{s_1} \\ S_2 \sreduces \strin{s_2}} 
+{\rsconcat{S_1}{S_2} \sreduces \strin{s_1 s_2}} % ???
+
+\inferrule[S-E-Replace]
+{S_1 \sreduces \strin{s_1} \\ S_2 \sreduces \strin{s_2} \\ \lsubst{r}{s_1}{s_2} = s} 
+{\sisubst{r}{S_1}{S_2} \sreduces \sistr{s}}
+
+\inferrule[S-E-Coerce-OK]
+{S \sreduces \strin{s} \\ s \in \lang{r}}
+{\coerce{r}{S} \sreduces \strin{s}}
+
+\inferrule[S-E-Coerce-Err]
+{S \sreduces \strin{s} \\ s \not \in \lang{r}}
+{\coerce{r}{S} \err}
+
+\end{mathpar}
+\caption{Big step semantics for our fragment of $\lambdas$.
+Error propagation rules are omitted.}
+\label{fig:eeval}
+\end{figure}
+
+In this section, we define a minimal typed lambda calculus with regular string types called $\lambda_{RS}$.
+This will serve as the source language for our translation to a calculus with only standard strings and regular expressions, defined in Sec.
+\ref{translation}.
+
+The syntax of $\lambda_{RS}$ is given in Figure \ref{fig:lamrs}, its static semantics are specified in Figure \ref{fig:etyping} and its evaluation semantics are specified in Figure \ref{fig:eeval}.
+
 The system has regular expression types $\stringin{r}$ whre $r$ is a regular expression. 
@@
 another string; for instance, $\lsubst{a|b}{a}{c} = c$.
-In order to compute the type resulting from aubstitution, we also need to compute
-the result of replacing one language with another inside a given language.
+In order to computer the type resulting from aubstitution, we also need to compute
+the result of replacing on language with another inside a given language.
 Finally, just for convienance, we provide a coerce operation. The introduction of
@@
 % Removes the < > from grammar productions.
-\renewcommand{\grammarlabel}[2]{#1\hfill#2}
-
-\begin{figure}[t]
-\begin{grammar}
-<r> ::= $\epsilon$ | $.$ | $a$ | $r \cdot r$ | $r + r$ | $r*$ \hfill $a \in \Sigma$
-
-\caption{Regular expressions over the alphabet $\Sigma$.}
-\label{fig:regex}
-\end{grammar}
-\end{figure}
-
-\begin{figure}[h]
-\begin{grammar}
-
-<$\psi$> ::=	$\psi \rightarrow \psi$			\hfill	source types					\alt
-$\stringin{r}$
-
-<S> ::= 
-      $\lambda x . e$ \hfill source terms \alt
-      $e e$ \alt
-      $\strin{s}$ \hfill $s \in \Sigma^{*}$ \alt
-      $\rsconcat{S}{S}$ \alt
-      $\sisubst{r}{S}{S}$ \alt
-      $\coerce{r}{S}$
-\caption{Syntax for the string sanitation fragment of our source language, $\lambdas$.}
-\end{grammar}
-\end{figure}
 
@@
 
-\ 
-\newpage
-\ 
-\newpage
-\
-
 \addtocounter{figure}{4}
@@
 %
-
-\newcommand{\sctx}{\Psi} % Context for external typing
-\newcommand{\tctx}{\Theta} % Context for internal typing
-
-\newcommand{\ereduces}{\Downarrow}
-
-\begin{figure}[t]\label{fig:lambdas}
-\small
-$\fbox{\inferrule{}{\sctx \vdash S : \psi}}$
-~~~~$\sctx ::= \emptyset \pipe \sctx, x : \psi$
-\begin{mathpar}
-%\inferrule[S-T-Abs]
-%{\sctx, x : T_1 \vdash e : T_2}
-%{\sctx \vdash \lambda x.e : T_1 \rightarrow T_2}
-%
-%\inferrule[S-T-App]
-%{\sctx \vdash e_1 : T' \\ \sctx \vdash e_2 : T' \rightarrow T}
-%{\sctx \vdash e_1 e_2 : T}
-%  
-\inferrule[S-T-Stringin-I]
-{s \in \lang{r}}
-{\sctx \vdash \strin{s} : \stringin{r}}
-
-\inferrule[S-T-Concat]
-{\sctx \vdash S_1 : \stringin{r_1} \\ \sctx \vdash S_2 : \stringin{r_2}}
-{\sctx \vdash \rsconcat{S_1}{S_2} : \stringin{r_1 \cdot r_2}}
-
-\inferrule[S-T-Replace]
-{\sctx \vdash S_1 : \stringin{r_1} \\ \sctx \vdash S_2 : \stringin{r_2} \\\\ \lreplace{r}{r_1}{r_2} = r'}
-{\sctx \vdash \sisubst{r}{S_1}{S_2} : \stringin{r'}}
-
-\inferrule[S-T-coerce]
-{\sctx \vdash S : \stringin{r'}}
-{\sctx \vdash \coerce{r}{S} : \stringin{r}}
-
-\end{mathpar}
-
-\caption{Typing rules for our fragment of $\lambdas$. 
-The typing context $\sctx$ is standard.}
-\label{fig:etyping}
-\end{figure}
-
-\begin{figure}[t]
-\small
-$\fbox{\inferrule{}{S \sreduces S}}$
-$\fbox{\inferrule{}{S \err}}$
-\begin{mathpar}
-%
-%\inferrule[S-E-Abs]                                                             
-%{ \ }                                                                           
-%{\lambda x.e \sreduces \lambda x.e}                                             
-%
-%\inferrule[S-E-App]
-%{ e_1 \sreduces e_1' \\ e_2 \sreduces e_2' }
-%{ e_1 e_2 \sreduces e_1' e_2'}
-%
-\inferrule[S-E-RStr]
-{ \ }
-{\strin{s} \sreduces \strin{s}}
-
-\inferrule[S-E-Concat]
-{S_1 \sreduces \strin{s_1} \\ S_2 \sreduces \strin{s_2}} 
-{\rsconcat{S_1}{S_2} \sreduces \strin{s_1 s_2}} % ???
-
-\inferrule[S-E-Replace]
-{S_1 \sreduces \strin{s_1} \\ S_2 \sreduces \strin{s_2} \\ \lsubst{r}{s_1}{s_2} = s} 
-{\sisubst{r}{S_1}{S_2} \sreduces \sistr{s}}
-
-\inferrule[S-E-Coerce-OK]
-{S \sreduces \strin{s} \\ s \in \lang{r}}
-{\coerce{r}{S} \sreduces \strin{s}}
-
-\inferrule[S-E-Coerce-Err]
-{S \sreduces \strin{s} \\ s \not \in \lang{r}}
-{\coerce{r}{S} \err}
-
-\end{mathpar}
-\caption{Big step semantics for our fragment of $\lambdas$.
-Error propagation rules are omitted.}
-\label{fig:eeval}
-\end{figure}
 
@@
 The typing context $\tctx$ is standard.}
-\label{fig:etyping}
+\label{fig:ityping}
 \end{figure}
@@
 
-The main theorem of this paper is Theorem \ref{thm:main}.
-Establishing this result requires an additional theorem (Theorem \ref{tcorrect}), which establishes a relationship between canonical forms for the string fragments of $\lambdas$ and $\lambdap$.
-
 \begin{thm}{Translation Correctness.} \label{thm:tcorrect}
@@
 
-\begin{thm}{Correctness of Input Sanitation for Translated Terms.}\label{thm:main}
+\begin{thm}{Correctness of Input Sanitation for Translated Terms.}
   If $\trden{S} = P$ and $\sctx \vdash S : \stringin{r}$ then either $P \err$ or $P \sreduces \tstr{s}$
@@
 
-\section{Implementation in {Atlang}}
+\section{Implementation as a Library in {Atlang}}\label{atlang}
 \begin{figure}
 \begin{lstlisting}
-#todo: why are these r's here?
-#TODO-nrf this is maybe kind-of right bue I think there's a nicer way to do it.
 @fn
 def sanitize(s : string_in[r'.*']):
-  return (s.replace(r'"', '&quot;') 
+  return (s.replace(r'"', '&quot;') # TODO: is this right?
            .replace(r'<', '&lt;')
@@
 
-In this section, we present an example atlang program using our extension together
-with our implementation of the extension in Atlang modulo ${\sf lreplace}$.
-Note that in addition to Figure \ref{fig:regex}, regular expressions in atlang
-also support complementation using the carot.
-
-TODO-nrf insert a pargraph or two which explains what's going on...
-
-\subsection{An Example}
-
-Figure \ref{fig:example} demosntrates the use of string types in atlang.
-The sanitize function takes an arbitrary string and returns a string without
-double quotes and left and right brackets. In this example, we use HTML
-escape sequences; many variants of SQL provide their own syntax
-for this as well.
-
-TODO-nrf line by line of example.
-
-\subsection{Extension Implementation}
-
-TODO-nrf line by line of extension implementation as ewll.
-
-An implementation of a similar extension as the one presented here is discussed in \cite{fulton12} and \cite{fulton13}.
-Our prior work only supports a special case of replacement where unsafe substrings are completely removed.
+\section{Related Work}\label{related}
+The input sanitation problem is well-understood.
+There exist a large number of techniques and technologies, proposed by both practitioners and researchers, for preventing injection-style attacks.
+In this section, we explain how our approach to the input sanitation problem differs from each of these approaches.
+More important than these differences, however, is our more general assertion that language extensibility is a promising approach toward consideration of security goals in programming lanugage design.
+
+Unlike \emph{frameworks and libraries} provided by languages such as Haskell and Ruby, our type system provides a \emph{static} guarantee that input is always properly sanitized before use.
+Doing so requires reasoning about the operations on regular languages corresponding to standard operations on strings; we are unaware of any production system which contains this form of reasoning.
+Therefore, even where frameworks and libraries provide a viable interface or wrapper around input sanitation, our approach is complementary because it ensures the correctness of the framework or library itself.
+Furthermore, our approach is more general than database abstraction layers because our mechanism is applicable to all forms of command injection (e.g.
+shell injection or remote file inclusion).
+
+A number of research languages provide static guarantees that a program is free of input sanitation vulnerabilities \cite{UrFlowOSDI10}. 
+Unlike this work, our solution to the input sanitation problem has a very low barrier to adoption; for instance, our implementation conservatively extends Python -- a popular language among web developers.
+We also believe our general approach is better-positioned for security, where continuously evolving threats might require frequent addition of new analyses; in these cases, the composability and generality of our approach is a substantial advantage.
+
+%\begin{itemize}
+%  \item Our system is a light-weight solution to a single class of sanitation vulnerabilities (e.g. we do not address Cross-Site Scripting).
+%  \item Our system is defined as a library in terms of an extensible type system, as opposed to a stand-alone language.
+Instead of introducing new technologies and methodologies for addressing security problems, we provide a light-weight static analysis which complements approaches developers already understand well.
+%  \item Our implementation of the translation is implemented in Python and shares its grammar.
+Since Python is a popular programming language among web developers, the barrier between our research and adopted technologies is lower than for greenfield security-oriented languages.
+%\end{itemize}
+
+We are also unaware of any extensible programming languages which emphasize applications to security concerns.
+
+Incorporating regular expressions into the type system is not novel.
+The XDuce system \cite{HosoyaVouillonPierce2000ICFP,HosoyaPierce2002} checks XML documents against schemas using regular expressions.
+Similarly, XHaskell \cite{xhaskell} focuses on XML documents.
+We differ from this and related work in at least three ways:
+\begin{itemize}
+  \item Our system is defined within an extensible type system.
+  \item We demonstrate that regular expression types are applicable to the web security domain, whereas previous work on regular expression types focused on XML schemas.
+  \item Although our static replacement operation is definable in some languages with regular expression types, we are the first to expose this operation and connect the semantics of regular language replacement with the semantics of string substitution via a type safety and compilation correctness argument.
+\end{itemize}
+
+In conclusion, our contribution is a type system, implemented within an extensible type system, for checking the correctness of input sanitation algorithms.
+\section{Conclusion}
+
+An implementation of a similar system as a type system extension in the programming language Ace is discussed in \cite{fulton12} and \cite{fulton13}.
+The implementation only supports a special case of replacement where unsafe substrings are simply stripped.
 In practice, most libraries and frameworks escape command sequences instead of stripping characters; therefore, we plan to extend our implementation with
-support for the replace operation as described in this paper.
+support for the replace operation as described in this paper..
 
@@
 %\\
-%$\psi$ & $\xlA{\psi}$ & \xA{\psi}\\
+%$\sigma$ & $\xlA{\sigma}$ & \xA{\sigma}\\
 %\hline

### 2014-08-13 sanitation-psp14/sanitation-psp14-short.tex

@@
 We support this by 1) specifying a  translation from $\lambda_{RS}$ to a language containing only strings and regular expressions, then, taking Python as such a language, 2) implement the type system together with the translation as a library using \texttt{atlang}, an extensible static type system for Python (being developed by the authors).% represents the most sophisticated application of this library to date.
+
 % Security-oriented programming languages do not enjoy wide adoption.
@@
 To avoid this problem, the program must \emph{sanitize} user input.
-For example, in this case, the developer (or, more often, a framework) might define a function \verb|sanitize| that prepends double quotes (and existing backslashes)  with a backslash, which SQL treats as safely.
+For example, in this case, the developer (or, more often, a framework) might define a function \verb|sanitize| that prepends double quotes (and existing backslashes)  with a backslash, which SQL treats safely.
 Guaranteeing that user input has already been sanitized before it is used to construct a command is challenging.
@@
 We write this as $\mathcal{L}\{r\}$.
-We take advantage of closure and decidability properties of regular languages to support a number of useful operations on values of such \emph{regular string types}.
+We take advantage of closure and decidability properties of regular languages to support a number of useful operations on values of such \emph{regular string types}, including \emph{replacement} of substrings matching a given regular expression.
 These make it possible to implement sanitation protocols like the one just described in an essentially conventional manner.
@@
 This approach is also \emph{complementary} to existing input sanitation techniques which use string replacement for input sanitation.
+% Our type system for regular string types, $\lambdas$, is defined as an exntension to a simply-typed lambda calculus.
+% Operations on strings are given types according to corresponding, type-level
+% computations on regular expressions. A concatenation operation is defined in 
+% order to demonstrate this principle, but the premier operation of our system
+% is \emph{replacement}, which captures a rich set of input sanitation techniques
+% including escaped characters and encodings. In addition to defining this system,
+% we present a correctness-preserving compilation of $\lambdas$ into an underlying
+% language with ordinary strings as a regular expression library.
+
+% In addition to this theoretical contribution, we also explain how our extension
+% may be implemented in the Ace Programming Language, which is itself an extension
+% of Python -- a popular web programming language.
+
+% Summarily, we present a simple type system extension which ensures the absence
+% of input sanitation vulnerabilities by statically checking the correctness and 
+% correct use of input sanitation algorithms. This approach is \emph{composable}
+% in the sense that it is a conservative extension. This approach is \emph{complementary}
+% to existing input sanitation techniques that use regular expressions for input sanitation.
 
@@
 another string; for instance, $\lsubst{a|b}{a}{c} = c$.
-In order to computer the type resulting from aubstitution, we also need to compute
-the result of replacing on language with another inside a given language.
-Finally, just for convienance, we provide a coerce operation. The introduction of
+In order to compute the type resulting from substitution, we also need to compute
+the result of replacing one language with another inside a given language.
+Finally, just for convenience, we provide a coerce operation. The introduction of
 coercion requires handling of runtime errors.
@@
 \begin{itemize}
-  \item Page 3 contains a definition of $\lambdas, \lambdap$ and the translation from $\lambdas$ to $\lambdap$.
+  \item Page 3 contains a definition of $\lambdas, \lambdap$ and the translation from $\lambdas$ to $\lambdap$. Grammar follows immediately at the top of page 4.
   \item In \S 2.1 we state some properties about regular expressions which are needed in our correctness proofs.
    \item In \S 2.2 we prove type safety for $\lambdap$ as well as both type safety and correctness for $\lambdas$.
-  \item In \S 2.3 we prove that translation preserves the correctness reesult about $\lambdas$.
+  \item In \S 2.3 we prove that translation preserves the correctness result about $\lambdas$.
 \end{itemize}
@@
 % Removes the < > from grammar productions.
+
+% \begin{figure}[h]
+% \begin{grammar}
+
+% <$\theta$> ::= $\theta \rightarrow \theta$ \hfill target types \alt
+% $\str$ \alt $\regex$
+
+% <P> ::= $\lambda x . e$ \hfill target terms \alt
+%   $e e$ \alt
+%   $\tstr{s}$ \alt
+%   $\rx{r}$ \alt
+%   $\tconcat{P}{P}$ \alt
+%   $\preplace{P}{P}{P}$ \alt
+%   $\tcheck{P}{P}$ 
+%   \end{grammar}
+% \caption{Syntax for the fragment of our target language, $\lambdap$, containing strings and statically constructed regular expressions.}
+% \label{fig:lcsSyntax}
+% \end{figure}
+
+%The $\lcs$ language gives static semantics for common regular expression library
+%functions. In this treatment, we include concatenation and filtering.
+%The ${\tt filter}$ function removes all instances of a regular expression in a string,
+%while concatenation ($+$) concatenates two strings.
+%
+%\subsection{Typing}
+%
+%The $\strin[r]$ type is parameterized by regular expressions; if $e:\strin[r]$,
+%then $e \in r$. Mapping from an arbitrary $\str$ to a $\strin[r]$ requires
+%defining an algorithm -- in terms of filter -- for converting a $\strin[.*]$
+%into a $\strin[r]$. The static semantics of the language defines the types of
+%operations on regular expressions in terms of well-understood properties about
+%regular lanugages; we recall these properties in section 3.
+%
+%\subsection{Dynamics}
+%
+%There are two evaluation judgements: $e:T \reduces e'$ and $e:T \treduces i$.
+%The $\reduces$ relation is between $\lcs$ expressions, while the $\treduces$
+%relation is a mapping from $\lcs$ expressions into internal language expressions
+%$i$ such that $i \ival$.
+%
+%Safety of the evaluation relation depends upon an injective mapping from $\lcs$ types info
+%internal language types. This relation, $h$, is defined below.
+%
+%\subsection{Type Safety}
+%
+%The type safety proof relies upon some assumptions about the type system and
+%dynamics of the internal language, as well as some properties of regular
+%languages.
+%
+%There must exist a translation from $\lcs$ types to the types of the internal
+%language. For the remainder of this paper, we call the type translation function
+%$h$.
+%
+%\begin{defn}[Type Translation Function $h$]
+%  The type translation function $h : Type \rightarrow IType$ is defined as follows:
+%  \begin{itemize}
+%    \item $\forall r. h(\strin[r]) = istr$
+%    \item $h(\str) = istr$
+%  \end{itemize}
+%\end{defn}
+%
+%Additionally, we assume that the internal language contains an implementation of strings,
+%together with operations for concatenation and filtering by regular expression.
+%
+%\begin{defn}[Types of internal values]
+%  Let $\tstrf{s}$ range over string literals and $r$ over regular expressions.
+%  Internal values are typed as follows:
+%\begin{itemize}
+%  \item If $e=\tstrf{s}$ then $e:istr$.
+%  \item If $e=\ifilter{r}{\tstrf{s}}$ then $e:istr$.
+%  \item If $e=\tstrf{s_1}+\tstrf{s_2}$ then $e:istr$.
+%\end{itemize}
+%\end{defn}
+%
+%For simplicity, we assume a fixed translation from
+%$\lcs$ regular expressions to regular expressions recognizable by the internal
+%language's regex library (in practice, a fixed translation is acceptable.)
+%To summarize, we assume an internal language containing 
+%a string type together with operations for string concatentation and filtering. We
+%expect closure over strings for both operations.
+%Finally, recall that ${\tt ifilter}$ is only needed for dynamic casts, which may be removed without
+%descreasing the expressivity or even usability of the language.
+%Finally, the semantics of the filter function are defined in terms of ${\tt rl\_filter}$,
+%which is a static version of ${\tt ifilter}$.
+%
+
+% I can't achieve the correct latout here. We should have something like:
+%       Lam_S Synax         Lam_P Syntax
+%       Lam_S Sematics      LAm_P Semantics
+%                Translation rules
+% Or, alternatively, we can flip reflect things so that the entire page
+% reads top-down (Lam_S syntax Lam_S smantics horizontally aligned w/ one
+% another).
+
+% Unfortunately, I can't get that to work. So, we're settling for something
+% slightly ugly, where the fitures appear in the incorrect order and the 
+% translation rules are on top.
+
+% The translations rules *just won't* go to the bottom of the page, so 
+% instead I am placing them at the top. Since the figures are out of order,
+% it's even more important that we get the number gith. therefore, the counter
+% is modified before and after this figure.
+
+\addtocounter{figure}{4}
+\begin{figure*}[t]\label{fig:trans}
+
+$\fbox{\inferrule{}{\trden{S} = P}}$
+\begin{mathpar}
+
+\inferrule[Tr-string]
+{ \ }
+{\trden{\strin{s}} = \tstr{s}}
+
+\inferrule[Tr-Concat]
+{ \trden{S_1} = P_1 \\ \trden{S_2} = P_2}
+{\trden{\rsconcat{S_1}{S_2}} = \tconcat{P_1}{P_2}}
+
+\inferrule[Tr-Subst]
+{ \trden{S_1} = P_1 \\ \trden{S_2} = P_2 }
+{\trden{ \sisubst{r}{S_1}{S_2} } = \tsubst{\rx{r}}{P_1}{P_2} }
+
+\inferrule[Tr-Coerce-Ok]
+{ S : \strin{r} \\ \lang{r'} \subseteq \lang{r}}
+{\trden{ \coerce{r'}{S} } = \tstr{s}}
+
+\inferrule[Tr-Coerce-NotOk]
+{\trden{S} = P \\ S : \strin{r} \\ \lang{r'} \not \subseteq \lang{r}}
+{\trden{\coerce{r'}{S}} = \tcheck{\rx{r'}}{P}}
+
+\end{mathpar}
+\caption{Translation from source terms (S) to target terms (P).
+The translation is type-directed in the Tr-Coerce cases.}
+\end{figure*}
+\addtocounter{figure}{-5}
+
+%
+% Begin typing/translation relation
+%
+
+\newpage
+
+\begin{figure}[t]\label{fig:lambdap}
+\small
+$\fbox{\inferrule{}{\tctx \vdash P : \theta}}$
+~~~~$\tctx ::= \emptyset \pipe \tctx, x : \theta$
+
+\begin{mathpar}
+%\inferrule[P-T-Abs]
+%{\tctx, x : T_1 \vdash e : T_2}
+%{\tctx \vdash \lambda x.e : T_1 \rightarrow T_2}
+%
+%\inferrule[P-T-App]
+%{\tctx \vdash e_1 : T \\ \tctx \vdash e_2 : T' \rightarrow T}
+%{\tctx \vdash e_1 e_2 : T}
+%
+\inferrule[P-T-String]
+{ \ }
+{\tctx \vdash \tstr{s} : \str}
+
+\inferrule[P-T-Regex]
+{ \ }
+{\tctx \vdash \rx{r} : \regex}
+
+\inferrule[P-T-Concat]
+{\tctx \vdash P_1 : \str \\ \tctx \vdash P_2 : \str}
+{\tctx \vdash \tconcat{P_1}{P_2} : \str}
+
+\inferrule[P-T-Replace]
+{\tctx \vdash P_1 : \regex \\ \tctx \vdash P_2 : \str \\ \tctx \vdash P_3 : \str }
+{\tctx \vdash \preplace{P_1}{P_2}{P_3} : \str}
+
+\inferrule[P-T-Check]
+{\tctx \vdash P_1 : \regex \\ \tctx \vdash P_2 : \str \\ }
+{\tctx \vdash \tcheck{P_1}{P_2} : \str}
+\end{mathpar}
+\caption{Typing rules for our fragment of $\lambdap$.
+The typing context $\tctx$ is standard.}
+\label{fig:ityping}
+\end{figure}
+
+\begin{figure}[t]
+\small
+$\fbox{\inferrule{}{P \treduces P}}$
+$\fbox{\inferrule{}{P \err}}$
+
+\begin{mathpar}
+
+%\inferrule[P-E-Abs]
+%{ \ }
+%{\lambda x.e \sreduces \lambda x.e}
+%
+%\inferrule[P-E-App]
+%{ e_1 \sreduces e_1' \\  e_2 \sreduces e_2' }
+%{ e_1 e_2 \sreduces e_1' e_2'}
+%
+  \inferrule[P-E-Str]
+{ \ }
+{\tstr{s} \treduces \tstr{s}}
+
+\inferrule[P-E-Rx]
+{ \ }
+{\rx{r} \treduces \rx{r}}
+
+\inferrule[P-E-Concat]
+{P_1 \treduces \tstr{s_1} \\ P_2 \treduces \tstr{s_2}} 
+{\tconcat{P_1}{P_2} \treduces \tstr{s_1 s_2}} % ???
+
+\inferrule[P-E-Replace]
+{P_1 \treduces \rx{r} \\ P_2 \treduces \tstr{s_2} \\ P_3 \treduces \tstr{s_3} \\ \lsubst{r}{s_2}{s_3} = s} 
+{\preplace{P_1}{P_2}{P_3} \treduces \tstr{s}}
+
+\inferrule[P-E-Check-OK]
+{P_1 \treduces \rx{r} \\ P_2 \treduces \strin{s} \\ s \in \lang{r}}
+{\tcheck{P_1}{P_2} \treduces \tstr{s}}
+
+\inferrule[P-E-Check-Err]
+{P_1 \treduces \rx{r} \\ P_2 \treduces \tstr{s} \\ s \not \in \lang{r}}
+{\tcheck{P_1}{P_2} \err}
+\end{mathpar}
+\caption{Big step semantics for our fragment of $\lambdap$.
+Error propagation rules are omitted.}
+\label{fig:ieval}
+\end{figure}
+
+\newpage
+\ 
+\newpage
+\ 
+\newpage
+\
+
+% Begin Grammar 
+% Removes the < > from grammar productions.
+\renewcommand{\grammarlabel}[2]{#1\hfill#2}
+
+% \begin{figure}[t]
+% \begin{grammar}
+% <r> ::= $\epsilon$ | $.$ | $a$ | $r \cdot r$ | $r + r$ | $r*$ \hfill $a \in \Sigma$
+
+% \caption{Regular expressions over the alphabet $\Sigma$.}
+% \label{fig:regex}
+% \end{grammar}
+% \end{figure}
+
+% \begin{figure}[h]
+% \begin{grammar}
+
+% <$\psi$> ::=	$\psi \rightarrow \psi$			\hfill	source types					\alt
+% $\stringin{r}$
+
+% <S> ::= 
+%       $\lambda x . e$ \hfill source terms \alt
+%       $e e$ \alt
+%       $\strin{s}$ \hfill $s \in \Sigma^{*}$ \alt
+%       $\rsconcat{S}{S}$ \alt
+%       $\sisubst{r}{S}{S}$ \alt
+%       $\coerce{r}{S}$
+% \caption{Syntax for the string sanitation fragment of our source language, $\lambdas$.}
+% \end{grammar}
+% \end{figure}
 
@@
 
-%The $\lcs$ language gives static semantics for common regular expression library
-%functions. In this treatment, we include concatenation and filtering.
-%The ${\tt filter}$ function removes all instances of a regular expression in a string,
-%while concatenation ($+$) concatenates two strings.
-%
-%\subsection{Typing}
-%
-%The $\strin[r]$ type is parameterized by regular expressions; if $e:\strin[r]$,
-%then $e \in r$. Mapping from an arbitrary $\str$ to a $\strin[r]$ requires
-%defining an algorithm -- in terms of filter -- for converting a $\strin[.*]$
-%into a $\strin[r]$. The static semantics of the language defines the types of
-%operations on regular expressions in terms of well-understood properties about
-%regular lanugages; we recall these properties in section 3.
-%
-%\subsection{Dynamics}
-%
-%There are two evaluation judgements: $e:T \reduces e'$ and $e:T \treduces i$.
-%The $\reduces$ relation is between $\lcs$ expressions, while the $\treduces$
-%relation is a mapping from $\lcs$ expressions into internal language expressions
-%$i$ such that $i \ival$.
-%
-%Safety of the evaluation relation depends upon an injective mapping from $\lcs$ types info
-%internal language types. This relation, $h$, is defined below.
-%
-%\subsection{Type Safety}
-%
-%The type safety proof relies upon some assumptions about the type system and
-%dynamics of the internal language, as well as some properties of regular
-%languages.
-%
-%There must exist a translation from $\lcs$ types to the types of the internal
-%language. For the remainder of this paper, we call the type translation function
-%$h$.
-%
-%\begin{defn}[Type Translation Function $h$]
-%  The type translation function $h : Type \rightarrow IType$ is defined as follows:
-%  \begin{itemize}
-%    \item $\forall r. h(\strin[r]) = istr$
-%    \item $h(\str) = istr$
-%  \end{itemize}
-%\end{defn}
-%
-%Additionally, we assume that the internal language contains an implementation of strings,
-%together with operations for concatenation and filtering by regular expression.
-%
-%\begin{defn}[Types of internal values]
-%  Let $\tstrf{s}$ range over string literals and $r$ over regular expressions.
-%  Internal values are typed as follows:
-%\begin{itemize}
-%  \item If $e=\tstrf{s}$ then $e:istr$.
-%  \item If $e=\ifilter{r}{\tstrf{s}}$ then $e:istr$.
-%  \item If $e=\tstrf{s_1}+\tstrf{s_2}$ then $e:istr$.
-%\end{itemize}
-%\end{defn}
-%
-%For simplicity, we assume a fixed translation from
-%$\lcs$ regular expressions to regular expressions recognizable by the internal
-%language's regex library (in practice, a fixed translation is acceptable.)
-%To summarize, we assume an internal language containing 
-%a string type together with operations for string concatentation and filtering. We
-%expect closure over strings for both operations.
-%Finally, recall that ${\tt ifilter}$ is only needed for dynamic casts, which may be removed without
-%descreasing the expressivity or even usability of the language.
-%Finally, the semantics of the filter function are defined in terms of ${\tt rl\_filter}$,
-%which is a static version of ${\tt ifilter}$.
-%
-
-% I can't achieve the correct latout here. We should have something like:
-%       Lam_S Synax         Lam_P Syntax
-%       Lam_S Sematics      LAm_P Semantics
-%                Translation rules
-% Or, alternatively, we can flip reflect things so that the entire page
-% reads top-down (Lam_S syntax Lam_S smantics horizontally aligned w/ one
-% another).
-
-% Unfortunately, I can't get that to work. So, we're settling for something
-% slightly ugly, where the fitures appear in the incorrect order and the 
-% translation rules are on top.
-
-% The translations rules *just won't* go to the bottom of the page, so 
-% instead I am placing them at the top. Since the figures are out of order,
-% it's even more important that we get the number gith. therefore, the counter
-% is modified before and after this figure.
-
-\addtocounter{figure}{4}
-\begin{figure*}\label{fig:trans}
-
-$\fbox{\inferrule{}{\trden{S} = P}}$
-\begin{mathpar}
-
-\inferrule[Tr-string]
-{ \ }
-{\trden{\strin{s}} = \tstr{s}}
-
-\inferrule[Tr-Concat]
-{ \trden{S_1} = P_1 \\ \trden{S_2} = P_2}
-{\trden{\rsconcat{S_1}{S_2}} = \tconcat{P_1}{P_2}}
-
-\inferrule[Tr-Subst]
-{ \trden{S_1} = P_1 \\ \trden{S_2} = P_2 }
-{\trden{ \sisubst{r}{S_1}{S_2} } = \tsubst{\rx{r}}{P_1}{P_2} }
-
-\inferrule[Tr-Coerce-Ok]
-{ S : \strin{r} \\ \lang{r'} \subseteq \lang{r}}
-{\trden{ \coerce{r'}{S} } = \tstr{s}}
-
-\inferrule[Tr-Coerce-NotOk]
-{\trden{S} = P \\ S : \strin{r} \\ \lang{r'} \not \subseteq \lang{r}}
-{\trden{\coerce{r'}{S}} = \tcheck{\rx{r'}}{P}}
-
-\end{mathpar}
-\caption{Translation from source terms (S) to target terms (P).
-The translation is type-directed in the Tr-Coerce cases.}
-\end{figure*}
-\addtocounter{figure}{-5}
-
-%
-% Begin typing/translation relation
-%
-
-\newpage
-
-\begin{figure}[t]\label{fig:lambdap}
-\small
-$\fbox{\inferrule{}{\tctx \vdash P : \theta}}$
-~~~~$\tctx ::= \emptyset \pipe \tctx, x : \theta$
-
-\begin{mathpar}
-%\inferrule[P-T-Abs]
-%{\tctx, x : T_1 \vdash e : T_2}
-%{\tctx \vdash \lambda x.e : T_1 \rightarrow T_2}
-%
-%\inferrule[P-T-App]
-%{\tctx \vdash e_1 : T \\ \tctx \vdash e_2 : T' \rightarrow T}
-%{\tctx \vdash e_1 e_2 : T}
-%
-\inferrule[P-T-String]
-{ \ }
-{\tctx \vdash \tstr{s} : \str}
-
-\inferrule[P-T-Regex]
-{ \ }
-{\tctx \vdash \rx{r} : \regex}
-
-\inferrule[P-T-Concat]
-{\tctx \vdash P_1 : \str \\ \tctx \vdash P_2 : \str}
-{\tctx \vdash \tconcat{P_1}{P_2} : \str}
-
-\inferrule[P-T-Replace]
-{\tctx \vdash P_1 : \regex \\ \tctx \vdash P_2 : \str \\ \tctx \vdash P_3 : \str }
-{\tctx \vdash \preplace{P_1}{P_2}{P_3} : \str}
-
-\inferrule[P-T-Check]
-{\tctx \vdash P_1 : \regex \\ \tctx \vdash P_2 : \str \\ }
-{\tctx \vdash \tcheck{P_1}{P_2} : \str}
-\end{mathpar}
-\caption{Typing rules for our fragment of $\lambdap$.
-The typing context $\tctx$ is standard.}
-\label{fig:ityping}
-\end{figure}
-
-\begin{figure}[t]
-\small
-$\fbox{\inferrule{}{P \treduces P}}$
-$\fbox{\inferrule{}{P \err}}$
-
-\begin{mathpar}
-
-%\inferrule[P-E-Abs]
-%{ \ }
-%{\lambda x.e \sreduces \lambda x.e}
-%
-%\inferrule[P-E-App]
-%{ e_1 \sreduces e_1' \\  e_2 \sreduces e_2' }
-%{ e_1 e_2 \sreduces e_1' e_2'}
-%
-  \inferrule[P-E-Str]
-{ \ }
-{\tstr{s} \treduces \tstr{s}}
-
-\inferrule[P-E-Rx]
-{ \ }
-{\rx{r} \treduces \rx{r}}
-
-\inferrule[P-E-Concat]
-{P_1 \treduces \tstr{s_1} \\ P_2 \treduces \tstr{s_2}} 
-{\tconcat{P_1}{P_2} \treduces \tstr{s_1 s_2}} % ???
-
-\inferrule[P-E-Replace]
-{P_1 \treduces \rx{r} \\ P_2 \treduces \tstr{s_2} \\ P_3 \treduces \tstr{s_3} \\ \lsubst{r}{s_2}{s_3} = s} 
-{\preplace{P_1}{P_2}{P_3} \treduces \tstr{s}}
-
-\inferrule[P-E-Check-OK]
-{P_1 \treduces \rx{r} \\ P_2 \treduces \strin{s} \\ s \in \lang{r}}
-{\tcheck{P_1}{P_2} \treduces \tstr{s}}
-
-\inferrule[P-E-Check-Err]
-{P_1 \treduces \rx{r} \\ P_2 \treduces \tstr{s} \\ s \not \in \lang{r}}
-{\tcheck{P_1}{P_2} \err}
-\end{mathpar}
-\caption{Big step semantics for our fragment of $\lambdap$.
-Error propagation rules are omitted.}
-\label{fig:ieval}
-\end{figure}
-
-\newpage
-\ 
-\newpage
-\ 
-\newpage
+\subsection{Definition of $\lambdas$}
+
+The $\lambdas$ system extends the simply-typed lambda calculus with regular
+expression types; an explanation of significant typing rule follows:
+\begin{itemize}
+  \item Rule S-T-Concat states that the type of concatenated strings is 
+    obtained by concatenating the regular expressions for each string.
+  \item Rule S-T-Replace defines the type of $\textsf{rreplace}$. The expression $\textsf{rreplace}$
+    evaluates to the result of substituting every string matching $r$ in $s_1$
+    with $s_2$. This operation corresponds to the \textsf{str_replace} function of PHP.
+    The type of these expressions is defined in terms of an extra-linguistic \textsf{lreplace}
+    function, which is defined later in this section.
+    The expression $\lreplace{r}{r_1}{r_2}$ is obtained by
+    starting with $r_1$, and replacing any subexpression matching $r$ with $r_2$.
+  \item Rule S-T-Coerce allows coercion; however, note that this coercion cannot
+    invalidate our safety property because we ensure appropraite checks are always
+    inserted wherever coercions are used.
+\end{itemize}
+
+\subsection{Definition of $\lambdap$}
+
+The system $\lambdap$ is a straight-forward extension of a simply typed lambda calculus
+with a string type and a regular expression tpye. We include two operations which are available in the regular expression
+library of any modern programming language. The check operation ensures that an expression
+recognizes a string, and the replace operation is string replacement.
+
+\subsection{Definition of Translation}
+
+The translation from $\lambdas$ to $\lambdap$ is defined in figure 5.
+The coercion cases are most interesting. If the safety of coercion in manifest in the
+types of the expressions, then no runtime check is inserted.
+If the safety of coercion is not manifest in the types, then a check is inserted.
+
+In practice, the type of a replacement rarely matches a specification.
+Therefore, it is convienant in an implementation to always insert the appropriate
+coercion, and then only raise type errors when an automatically inserted coercion
+actually requires the insertion of a runtime check. Alternatively, this policy may 
+be codified in the type system itself using subtyping \cite{fulton12}.
 
@@
 
-%\begin{thm}{Correctness of Input Sanitation (CIS) for $\lambdas$.}\label{thm:scorrect}
-%  If $\sctx \vdash S : \stringin{r}$ then either (a) $S \sreduces \rstr{s}$ for $s \in \lang{r}$
-%  or (b) $S \err$.
-%\end{thm}
-
-\begin{thm} 
-  Let $P$ be a term in the target language. If $\tctx \vdash P : \theta$ 
-  then $P \sreduces P'$ and $\tctx \vdash P' : \theta$, or else $P \err$.
-\end{thm}
-
 \subsection{Translation Correctness}
+
+The main theorem of this paper is Theorem \ref{thm:main},
+which establishes that Theorem \ref{thm:scorrect} is preserved under translation
+into the target language $\lambdap$.
+
+Establishing this result requires an additional theorem establishing a relationship between canonical forms for the string fragments of $\lambdas$ and $\lambdap$.
 
@@
 
-\section{Implementation as a Library in {Atlang}}\label{atlang}
-\begin{figure}
+\section{Implementation in {Atlang}}\label{atlang}
+\begin{figure}\label{fig:example}
 \begin{lstlisting}
@@
 def sanitize(s : string_in[r'.*']):
-  return (s.replace(r'"', '&quot;') # TODO: is this right?
+  return (s.replace(r'"', '&quot;') 
            .replace(r'<', '&lt;')
@@
 \end{figure}
-wwwww
-Here, we use the argument annotation syntax introduced in Python 3 (a similar syntax is available for Python 2.6+, not shown).
-
-\section{Related Work}\label{related}
+
+\subsection{An Example}
+
+Figure \ref{fig:example} demosntrates the use of string types in atlang.
+The sanitize function takes an arbitrary string and returns a string without
+double quotes or left and right brackets. In this example, we use HTML
+escape sequences.
+
+The main function receives user input and passes this input to a 
+sanitize function, which replaces all double quotes and brackets with HTML
+escape sequences.
+
+The result of applying sanitize to input is appended to two functions which
+construct a safe query and safe output.
+The arguments to the result and output construction functions constitute \emph{specifications}.
+In the case of \textsf{results_query}, this specification ensures that user input is always interpreted
+as a string literal by the SQL server.
+In the case of \textsf{results_div}, this specification ensures that user input does 
+not contain any HTML tags, which is a conservative but effective policy for preventing
+XSS attacks.
+
+Note that ${\sf input}$ does not actually meet these specifications without additional machinery.
+The type of ${\sf input}$ is quite large
+and does not actually equal the specified domains of the query or output construction
+methods. This mismatch is common -- in fact, nearly universal. Therefore, 
+our implementation includes a simple subtyping relation between regular expression
+types. 
+This subtyping relation is justified theoretically by the fact that language inclusion is
+decidable; see \cite{fulton12} for a formal definition of the subtyping relation.
+Additionally, our extension remains composable because subtyping is defined on a type-by-type basis;
+see \cite{fulton13} for a discussion of subtyping in Atlang (referred to there as Ace).
+
+\subsection{Implementation of the Regular Expression Type}
+
+We implemented a variation on the type system presented in this paper. The only
+significant difference is that we only support replacements where $s_2$ is the empty
+string. Therefore, our implementation respects the system presented in section 2
+only modulo the definition of \textsf{lreplace}.
+
+Atlang translates programs using type definitions, which may extend both the
+static and dynamic semantics of the language. New types are defined as Python
+classes; figure \ref{fig:impl} contains the source code of our implementation.
+
+The \textsf{string_in} type has an indexing regular expression \texttt{idx}.
+Our translation is defined by the \textsf{trans_} methods while the \textsf{syn_}
+methods define our type checker. Atlang defers type checking and translation
+to these methods whenever an expression of type \textsf{string_in} is encountered.
+
+\subsection{Related Work and Alternative Approaches}
+
 The input sanitation problem is well-understood.
@@
 
+The Wyvern programming language provides a general framework for composing language extensions \cite{tsl}\cite{wyvern}.
+Our work identifies one particular extension, and is therefore complementary to Wyvern and related
+work on extensible programming languages.
 We are also unaware of any extensible programming languages which emphasize applications to security concerns.
@@
 In conclusion, our contribution is a type system, implemented within an extensible type system, for checking the correctness of input sanitation algorithms.
-\section{Conclusion}
-
-An implementation of a similar system as a type system extension in the programming language Ace is discussed in \cite{fulton12} and \cite{fulton13}.
-The implementation only supports a special case of replacement where unsafe substrings are simply stripped.
-In practice, most libraries and frameworks escape command sequences instead of stripping characters; therefore, we plan to extend our implementation with
-support for the replace operation as described in this paper..
 

### 2014-08-13 sanitation-psp14/sanitation-psp14-short.tex
minor
Small presentation changes
use e for the source language terms and \iota for the target language terms

use \sigma and \tau for their types

use semicolons to separate arguments to operators instead of commas

lam_S -> lam_RS

@@
 %%
-\newcommand{\lambdas}{\lambda_S}
+\newcommand{\lambdas}{\lambda_{RS}}
 \newcommand{\lambdap}{\lambda_P}
@@
 % Source language terms.
-\newcommand{\sisubst}[3]{{\sf rreplace}[#1](#2,#3)} \newcommand{\rreplace}[3]{{\sf rreplace}[#1](#2,#3)} % lots of legacy naming around these parts...
+\newcommand{\sisubst}[3]{{\sf rreplace}[#1](#2;#3)} \newcommand{\rreplace}[3]{{\sf rreplace}[#1](#2;#3)} % lots of legacy naming around these parts...
 
@@
 \newcommand{\strin}[1]{\sistr{#1}}
-\newcommand{\rsconcat}[2]{{\sf rconcat}(#1,#2)} \newcommand{\rconcat}[2]{{\sf rconcat}(#1,#2)} % lots of legact naming around these parts..
+\newcommand{\rsconcat}[2]{{\sf rconcat}(#1;#2)} \newcommand{\rconcat}[2]{{\sf rconcat}(#1;#2)} % lots of legact naming around these parts..
 
@@
 % target language terms.
-\newcommand{\tsubst}[3]{{\sf replace}(#1,#2,#3)} \newcommand{\metareplace}[3]{{\sf replace}(#1,#2,#3)} % TODO-nrf rename the commands. Lots of legacy naming around these parts...
-
-\newcommand{\tcheck}[2]{{\sf check}(#1, #2)}
+\newcommand{\tsubst}[3]{{\sf replace}(#1;#2;#3)} \newcommand{\metareplace}[3]{{\sf replace}(#1;#2;#3)} % TODO-nrf rename the commands. Lots of legacy naming around these parts...
+
+\newcommand{\tcheck}[2]{{\sf check}(#1; #2)}
 \renewcommand{\tstr}[1]{{{\sf str}[#1]}}
-\newcommand{\preplace}[3]{{\sf preplace}(#1,#2,#3)}
-\newcommand{\tconcat}[2]{{\sf concat}(#1,#2)} \newcommand{\concat}[2]{{\sf concat}(#1,#2)} % lots of legacy naming around these parts...
+\newcommand{\preplace}[3]{{\sf preplace}(#1;#2;#3)}
+\newcommand{\tconcat}[2]{{\sf concat}(#1;#2)} \newcommand{\concat}[2]{{\sf concat}(#1;#2)} % lots of legacy naming around these parts...
 
@@
 % Meta-theoretic functions
-\newcommand{\lsubst}[3]{{\tt lsubst}(#1,#2,#3)} % This used to render lreplace(...) so there're probably mistkes wherever this command was used now.
-\newcommand{\lreplace}[3]{{\tt lreplace}(#1, #2, #3)}
+\newcommand{\lsubst}[3]{{\tt lsubst}(#1;#2;#3)} % This used to render lreplace(...) so there're probably mistkes wherever this command was used now.
+\newcommand{\lreplace}[3]{{\tt lreplace}(#1; #2; #3)}
 
@@
 \inferrule[S-T-Concat]
-{\sctx \vdash S_1 : \stringin{r_1} \\ \sctx \vdash S_2 : \stringin{r_2}}
-{\sctx \vdash \rsconcat{S_1}{S_2} : \stringin{r_1 \cdot r_2}}
+{\sctx \vdash e_1 : \stringin{r_1} \\ \sctx \vdash e_2 : \stringin{r_2}}
+{\sctx \vdash \rsconcat{e_1}{e_2} : \stringin{r_1 \cdot r_2}}
 
 \inferrule[S-T-Replace]
-{\sctx \vdash S_1 : \stringin{r_1} \\ \sctx \vdash S_2 : \stringin{r_2} \\\\ \lreplace{r}{r_1}{r_2} = r'}
-{\sctx \vdash \sisubst{r}{S_1}{S_2} : \stringin{r'}}
+{\sctx \vdash e_1 : \stringin{r_1} \\ \sctx \vdash e_2 : \stringin{r_2} \\\\ \lreplace{r}{r_1}{r_2} = r'}
+{\sctx \vdash \sisubst{r}{e_1}{e_2} : \stringin{r'}}
 
 \inferrule[S-T-coerce]
-{\sctx \vdash S : \stringin{r'}}
-{\sctx \vdash \coerce{r}{S} : \stringin{r}}
+{\sctx \vdash e : \stringin{r'}}
+{\sctx \vdash \coerce{r}{e} : \stringin{r}}
 
@@
 \inferrule[S-E-Concat]
-{S_1 \sreduces \strin{s_1} \\ S_2 \sreduces \strin{s_2}} 
-{\rsconcat{S_1}{S_2} \sreduces \strin{s_1 s_2}} % ???
+{e_1 \sreduces \strin{s_1} \\ e_2 \sreduces \strin{s_2}} 
+{\rsconcat{e_1}{e_2} \sreduces \strin{s_1 s_2}} % ???
 
 \inferrule[S-E-Replace]
-{S_1 \sreduces \strin{s_1} \\ S_2 \sreduces \strin{s_2} \\ \lsubst{r}{s_1}{s_2} = s} 
-{\sisubst{r}{S_1}{S_2} \sreduces \sistr{s}}
+{e_1 \sreduces \strin{e_1} \\ e_2 \sreduces \strin{s_2} \\ \lsubst{r}{s_1}{s_2} = s} 
+{\sisubst{r}{e_1}{e_2} \sreduces \sistr{s}}
 
@@
 {\coerce{r}{S} \sreduces \strin{s}}
-
-\inferrule[S-E-Coerce-Err]
-{S \sreduces \strin{s} \\ s \not \in \lang{r}}
-{\coerce{r}{S} \err}
 
@@
 \inferrule[Tr-Concat]
-{ \trden{S_1} = P_1 \\ \trden{S_2} = P_2}
-{\trden{\rsconcat{S_1}{S_2}} = \tconcat{P_1}{P_2}}
+{ \trden{e_1} = \iota_1 \\ \trden{e_2} = \iota_2}
+{\trden{\rsconcat{e_1}{e_2}} = \tconcat{\iota_1}{\iota_2}}
 
 \inferrule[Tr-Subst]
-{ \trden{S_1} = P_1 \\ \trden{S_2} = P_2 }
-{\trden{ \sisubst{r}{S_1}{S_2} } = \tsubst{\rx{r}}{P_1}{P_2} }
+{ \trden{e_1} = \iota_1 \\ \trden{e_2} = \iota_2 }
+{\trden{ \sisubst{r}{e_1}{e_2} } = \tsubst{\rx{r}}{\iota_1}{\iota_2} }
 
 \inferrule[Tr-Coerce-Ok]
-{ S : \strin{r} \\ \lang{r'} \subseteq \lang{r}}
-{\trden{ \coerce{r'}{S} } = \tstr{s}}
+{ e : \strin{r} \\ \lang{r'} \subseteq \lang{r}}
+{\trden{ \coerce{r'}{e} } = \tstr{s}}
 
 \inferrule[Tr-Coerce-NotOk]
-{\trden{S} = P \\ S : \strin{r} \\ \lang{r'} \not \subseteq \lang{r}}
-{\trden{\coerce{r'}{S}} = \tcheck{\rx{r'}}{P}}
+{\trden{e} = \iota \\ e : \strin{r} \\ \lang{r'} \not \subseteq \lang{r}}
+{\trden{\coerce{r'}{e}} = \tcheck{\rx{r'}}{\iota}}
 
@@
 \inferrule[P-T-Concat]
-{\tctx \vdash P_1 : \str \\ \tctx \vdash P_2 : \str}
-{\tctx \vdash \tconcat{P_1}{P_2} : \str}
+{\tctx \vdash \iota_1 : \str \\ \tctx \vdash \iota_2 : \str}
+{\tctx \vdash \tconcat{\iota_1}{\iota_2} : \str}
 
 \inferrule[P-T-Replace]
-{\tctx \vdash P_1 : \regex \\ \tctx \vdash P_2 : \str \\ \tctx \vdash P_3 : \str }
-{\tctx \vdash \preplace{P_1}{P_2}{P_3} : \str}
+{\tctx \vdash \iota_1 : \regex \\ \tctx \vdash \iota_2 : \str \\ \tctx \vdash \iota_3 : \str }
+{\tctx \vdash \preplace{\iota_1}{\iota_2}{\iota_3} : \str}
 
 \inferrule[P-T-Check]
-{\tctx \vdash P_1 : \regex \\ \tctx \vdash P_2 : \str \\ }
-{\tctx \vdash \tcheck{P_1}{P_2} : \str}
+{\tctx \vdash \iota_1 : \regex \\ \tctx \vdash \iota_2 : \str \\ }
+{\tctx \vdash \tcheck{\iota_1}{\iota_2} : \str}
 \end{mathpar}
@@
 \inferrule[P-E-Concat]
-{P_1 \treduces \tstr{s_1} \\ P_2 \treduces \tstr{s_2}} 
-{\tconcat{P_1}{P_2} \treduces \tstr{s_1 s_2}} % ???
+{\iota_1 \treduces \tstr{s_1} \\ \iota_2 \treduces \tstr{s_2}} 
+{\tconcat{\iota_1}{\iota_2} \treduces \tstr{s_1 s_2}} % ???
 
 \inferrule[P-E-Replace]
-{P_1 \treduces \rx{r} \\ P_2 \treduces \tstr{s_2} \\ P_3 \treduces \tstr{s_3} \\ \lsubst{r}{s_2}{s_3} = s} 
-{\preplace{P_1}{P_2}{P_3} \treduces \tstr{s}}
+{\iota_1 \treduces \rx{r} \\ \iota_2 \treduces \tstr{s_2} \\ \iota_3 \treduces \tstr{s_3} \\ \lsubst{r}{s_2}{s_3} = s} 
+{\preplace{\iota_1}{\iota_2}{\iota_3} \treduces \tstr{s}}
 
 \inferrule[P-E-Check-OK]
-{P_1 \treduces \rx{r} \\ P_2 \treduces \strin{s} \\ s \in \lang{r}}
-{\tcheck{P_1}{P_2} \treduces \tstr{s}}
+{\iota_1 \treduces \rx{r} \\ \iota_2 \treduces \strin{s} \\ s \in \lang{r}}
+{\tcheck{\iota_1}{\iota_2} \treduces \tstr{s}}
 
 \inferrule[P-E-Check-Err]
-{P_1 \treduces \rx{r} \\ P_2 \treduces \tstr{s} \\ s \not \in \lang{r}}
-{\tcheck{P_1}{P_2} \err}
+{\iota_1 \treduces \rx{r} \\ \iota_2 \treduces \tstr{s} \\ s \not \in \lang{r}}
+{\tcheck{\iota_1}{\iota_2} \err}
 \end{mathpar}
@@
 \begin{lem}
-  If $\sctx \vdash S : \stringin{r}$ then $r$ is a well-formed regular expression.
+  If $\sctx \vdash e : \stringin{r}$ then $r$ is a well-formed regular expression.
 \end{lem}
@@
 \begin{lem}
-If $\tctx \vdash P : \regex$ then $P \treduces \rx{r}$ such that $r$ is a well-formed regular expression. 
+If $\tctx \vdash \iota : \regex$ then $\iota \treduces \rx{r}$ such that $r$ is a well-formed regular expression. 
 \end{lem}
@@
 \begin{thm} 
-  Let $P$ be a term in the target language. If $\tctx \vdash P : \theta$ 
-  then $P \sreduces P'$ and $\tctx \vdash P' : \theta$, or else $P \err$.
+  Let $\iota$ be a term in the target language. If $\tctx \vdash \iota : \tau$ 
+  then $\iota \sreduces \iota'$ and $\tctx \vdash \iota' : \tau$.
 \end{thm}
@@
 \begin{thm}{Type Safety for the String Fragment of $\lambdas$.} \label{thm:scorrect}
-  Let $S$ be a term in the source language. If $\sctx \vdash S : \stringin{r}$ 
-  then $S \sreduces \rstr{s}$ and $\sctx \vdash \rstr{s} : \stringin{r}$; or else $S \err$.
+  Let $S$ be a term in the source language. If $\sctx \vdash e : \stringin{r}$ 
+  then $e \sreduces \rstr{s}$ and $\sctx \vdash \rstr{s} : \stringin{r}$.
 \end{thm}
 \begin{proof}
-By induction on the typing relation, where (a) case holds by lemma \ref{thm:regexprops} in the S-T-Concat case and lemma \ref{thm:substcorrespondence} in the S-T-Replace case..
-The (b) cases hold by unstated, but standard, error propagation rules.
+By induction on the typing relation.
+The S-T-Concat case requires Lemma \ref{thm:regexprops} 
+and the S-T-Replace case appeals to Lemma \ref{thm:substcorrespondence}.
 \end{proof}
@@
 \begin{thm}{Correctness of Input Sanitation for $\lambdas$.}\label{thm:scorrect}
-  If  $\sctx \vdash S : \stringin{r}$ and $S \sreduces \rstr{s}$ then $s \in \lang{r}$.
+  If  $\sctx \vdash e : \stringin{r}$ and $e \sreduces \rstr{s}$ then $e \in \lang{r}$.
 \end{thm}
@@
 \begin{thm}{Translation Correctness.} \label{thm:tcorrect}
-If $\sctx \vdash S : \stringin{r}$ then there exists a $P$ such that $\trden{S} = P$ and either: 
-(a) $P \treduces \tstr{s}$ and $S \sreduces \strin{s}$, or
-(b) $P \err$ and $S \err$.
+If $\sctx \vdash e : \stringin{r}$ then there exists a $\iota$ such that $\trden{e} = \iota$ such that 
+$iota \treduces \tstr{s}$ and $e \sreduces \strin{s}$, or
 \end{thm}
@@
 \begin{proof}
-The proof proceeds by induction on the typing relation for $S$ and an appropriate
-choice of $P$; in each case, the choice is obvious. The subcases (a) proceed
-by inversion and appeals to our type safety theorems as well as the induction hypothesis. The subcases (b) proceed
-by the standard error propagation rules omitted for space.
-Throughout the proof, properties from the closure lemma for regular languages
-are necessary.
+The proof proceeds by induction on the typing relation for $e$ and an appropriate
+choice of $\iota$; in each case, the choice is obvious and the rest of the proof
+proceeds by our type safety theorems as well as an appeal to the induction hypothesis.
 \end{proof}
@@
 \begin{thm}{Correctness of Input Sanitation for Translated Terms.}\label{thm:main}
-  If $\trden{S} = P$ and $\sctx \vdash S : \stringin{r}$ then either $P \err$ or $P \sreduces \tstr{s}$
+  If $\trden{e} = \iota$ and $\sctx \vdash e : \stringin{r}$ then $\iota \sreduces \tstr{s}$
   for $s \in \lang{r}$.
@@
 \begin{proof}
-  By theorem \ref{thm:tcorrect}, $P \sreduces \tstr{s}$ implies that $S \sreduces \strin{s}$.
-  By theorem \ref{thm:scorrect}, the above property together with the assumption that $S$ is well-typed implies that $s \in \lang{r}$.
+  By theorem \ref{thm:tcorrect}, $\iota \sreduces \tstr{s}$ implies that $e \sreduces \strin{s}$.
+  By theorem \ref{thm:scorrect}, the above property together with the assumption that $e$ is well-typed implies that $s \in \lang{r}$.
 \end{proof}

### 2014-08-13 sanitation-psp14/sanitation-psp14-short.tex
two typos from previous commit.

@@
 \begin{thm}{Correctness of Input Sanitation for $\lambdas$.}\label{thm:scorrect}
-  If  $\sctx \vdash e : \stringin{r}$ and $e \sreduces \rstr{s}$ then $e \in \lang{r}$.
+  If  $\sctx \vdash e : \stringin{r}$ and $e \sreduces \rstr{s}$ then $s \in \lang{r}$.
 \end{thm}
@@
 \begin{thm}{Translation Correctness.} \label{thm:tcorrect}
-If $\sctx \vdash e : \stringin{r}$ then there exists a $\iota$ such that $\trden{e} = \iota$ such that 
-$iota \treduces \tstr{s}$ and $e \sreduces \strin{s}$, or
+If $\sctx \vdash e : \stringin{r}$ then there exists a $\iota$ such that $\trden{e} = \iota$, 
+$\iota \treduces \tstr{s}$, and $e \sreduces \strin{s}$, or
 \end{thm}

### 2014-08-13 sanitation-psp14/sanitation-psp14-short.tex
Added check rules; lam_rs and lam_p are now full systems.
still need to proof read

@@
 \newcommand{\val}{{\sf val}}
+
+\newcommand{\rcheck}[4]{ {\sf rcheck}[#1](#2;#3;#4) }
 
@@
       $\sisubst{r}{e}{e}$ \alt
-      $\coerce{r}{e}$
+      $\coerce{r}{e}{x.e}{e}$
 \caption{Syntax of $\lambda_{RS}$.}\label{fig:lamrs}
@@
 \begin{mathpar}
-%\inferrule[S-T-Abs]
-%{\sctx, x : T_1 \vdash e : T_2}
-%{\sctx \vdash \lambda x.e : T_1 \rightarrow T_2}
-%
-%\inferrule[S-T-App]
-%{\sctx \vdash e_1 : T' \\ \sctx \vdash e_2 : T' \rightarrow T}
-%{\sctx \vdash e_1 e_2 : T}
-%  
+\inferrule[S-T-Abs]
+{\sctx, x : \sigma_1 \vdash e : \sigma_2}
+{\sctx \vdash \lambda x.e : \sigma_1 \rightarrow \sigma_2}
+
+\inferrule[S-T-App]
+{\sctx \vdash e_1 : \sigma_1 \\ \sctx \vdash e_2 : \sigma_1 \rightarrow \sigma}
+{\sctx \vdash e_1 e_2 : \sigma}
+
 \inferrule[S-T-Stringin-I]
@@
 
-\inferrule[S-T-coerce]
-{\sctx \vdash e : \stringin{r'}}
+\inferrule[S-T-SafeCoerce]
+{\sctx \vdash e : \stringin{r'} \\ \lang{r'} \subseteq \lang{r}}
 {\sctx \vdash \coerce{r}{e} : \stringin{r}}
 
+\inferrule[S-T-Check]
+{\sctx \vdash e_0 : \stringin{r_0} \\ \sctx, x:\stringin{r} \vdash e_1 : \sigma \\ \sctx \vdash e_2 : \sigma}
+{\sctx \vdash \rcheck{r}{e_0}{x.e_1}{e_2} : \sigma}
+
 \end{mathpar}
 
-\caption{Typing rules for our fragment of $\lambdas$. 
+\caption{Typing rules for $\lambdas$. 
 The typing context $\sctx$ is standard.}
@@
 \begin{mathpar}
-%
-%\inferrule[S-E-Abs]                                                             
-%{ \ }                                                                           
-%{\lambda x.e \sreduces \lambda x.e}                                             
-%
-%\inferrule[S-E-App]
-%{ e_1 \sreduces e_1' \\ e_2 \sreduces e_2' }
-%{ e_1 e_2 \sreduces e_1' e_2'}
-%
+\inferrule[S-E-Abs]                                                             
+{ \ }                                                                           
+{\lambda x.e \sreduces \lambda x.e}
+
+\inferrule[S-E-App]
+{ e_1 \sreduces e_1' \\ e_2 \sreduces e_2' }
+{ e_1 e_2 \sreduces e_1' e_2'}
+
 \inferrule[S-E-RStr]
@@
 
-\inferrule[S-E-Coerce-OK]
-{S \sreduces \strin{s} \\ s \in \lang{r}}
-{\coerce{r}{S} \sreduces \strin{s}}
+\inferrule[S-E-SafeCoerce]
+{e \sreduces \strin{s}}
+{\coerce{r}{e} \sreduces \strin{s}}
+
+\inferrule[S-E-Check-Ok]
+{e \sreduces \strin{s} \\ s \in \lang{r}}
+{
+  \rcheck{r}{e}{x.e_1}{e_2} \sreduces [\strin{s} / x]e_1
+}
+
+\inferrule[S-E-Check-NotOk]
+{
+  e \sreduces \strin{s} \\ s \not \in \lang{r}
+}
+{
+  \rcheck{r}{e}{x.e_1}{e_2} \sreduces e_2
+}
 
 \end{mathpar}
-\caption{Big step semantics for our fragment of $\lambdas$.
+\caption{Big step semantics for $\lambdas$.
 Error propagation rules are omitted.}
@@
 \begin{mathpar}
-%\inferrule[P-T-Abs]
-%{\tctx, x : T_1 \vdash e : T_2}
-%{\tctx \vdash \lambda x.e : T_1 \rightarrow T_2}
-%
-%\inferrule[P-T-App]
-%{\tctx \vdash e_1 : T \\ \tctx \vdash e_2 : T' \rightarrow T}
-%{\tctx \vdash e_1 e_2 : T}
-%
+\inferrule[P-T-Abs]
+{\tctx, x : \iota_1 \vdash e : \iota_2}
+{\tctx \vdash \lambda x.e : \iota_1 \rightarrow \iota_2}
+
+\inferrule[P-T-App]
+{\tctx \vdash e_1 : \iota_1 \\ \tctx \vdash e_2 : \iota_1 \rightarrow \iota}
+{\tctx \vdash e_1 e_2 : \iota}
+
 \inferrule[P-T-String]
@@
 \end{mathpar}
-\caption{Typing rules for our fragment of $\lambdap$.
+\caption{Typing rules for $\lambdap$.
 The typing context $\tctx$ is standard.}
@@
 \begin{mathpar}
-
-%\inferrule[P-E-Abs]
-%{ \ }
-%{\lambda x.e \sreduces \lambda x.e}
-%
-%\inferrule[P-E-App]
-%{ e_1 \sreduces e_1' \\  e_2 \sreduces e_2' }
-%{ e_1 e_2 \sreduces e_1' e_2'}
-%
+\inferrule[P-E-Abs]
+{ \ }
+{\lambda x.e \sreduces \lambda x.e}
+
+\inferrule[P-E-App]
+{ e_1 \sreduces e_1' \\  e_2 \sreduces e_2' }
+{ e_1 e_2 \sreduces e_1' e_2'}
+
   \inferrule[P-E-Str]
@@
 \end{mathpar}
-\caption{Big step semantics for our fragment of $\lambdap$.
+\caption{Big step semantics for of $\lambdap$.
 Error propagation rules are omitted.}
@@
     starting with $r_1$, and replacing any subexpression matching $r$ with $r_2$.
-  \item Rule S-T-Coerce allows coercion; however, note that this coercion cannot
-    invalidate our safety property because we ensure appropraite checks are always
-    inserted wherever coercions are used.
+  \item Rule S-T-SafeCoerce allows only safe coercions between string types by
+    expoiting the decidability of language inclusion.
 \end{itemize}
@@
 
-Safety for the string fragment is more involved, because it involves validating that
+Safety for the string fragment of $\lambdas$ is more involved because it involves validating that
 the type system's definition is justified by our theorems about regular languages.
+We state the theorem for the string fragment only to demontrate this point;
+full type safety for $\lambdas$ follows easily from safety for the string fagment.
 Note that type safety does \emph{not} guarantee that strings of a type are in the
@@
 If $\sctx \vdash e : \stringin{r}$ then there exists a $\iota$ such that $\trden{e} = \iota$, 
-$\iota \treduces \tstr{s}$, and $e \sreduces \strin{s}$, or
+$\iota \treduces \tstr{s}$, and $e \sreduces \strin{s}$.
 \end{thm}

### 2014-08-13 sanitation-psp14/sanitation-psp14-short.tex
A few minor edits.
check
Updates

--- 2014-08-13 sanitation-psp14/sanitation-psp14-short.tex
@@
 %%
-\newcommand{\lambdas}{\lambda_{RS}}
+\newcommand{\lambdas}{\lambda_S}
 \newcommand{\lambdap}{\lambda_P}
@@
 % Source language terms.
-\newcommand{\sisubst}[3]{{\sf rreplace}[#1](#2;#3)} \newcommand{\rreplace}[3]{{\sf rreplace}[#1](#2;#3)} % lots of legacy naming around these parts...
+\newcommand{\sisubst}[3]{{\sf rreplace}[#1](#2,#3)} \newcommand{\rreplace}[3]{{\sf rreplace}[#1](#2,#3)} % lots of legacy naming around these parts...
 
@@
 \newcommand{\strin}[1]{\sistr{#1}}
-\newcommand{\rsconcat}[2]{{\sf rconcat}(#1;#2)} \newcommand{\rconcat}[2]{{\sf rconcat}(#1;#2)} % lots of legact naming around these parts..
+\newcommand{\rsconcat}[2]{{\sf rconcat}(#1,#2)} \newcommand{\rconcat}[2]{{\sf rconcat}(#1,#2)} % lots of legact naming around these parts..
 
@@
 % target language terms.
-\newcommand{\tsubst}[3]{{\sf replace}(#1;#2;#3)} \newcommand{\metareplace}[3]{{\sf replace}(#1;#2;#3)} % TODO-nrf rename the commands. Lots of legacy naming around these parts...
-
-\newcommand{\tcheck}[2]{{\sf check}(#1; #2)}
+\newcommand{\tsubst}[3]{{\sf replace}(#1,#2,#3)} \newcommand{\metareplace}[3]{{\sf replace}(#1,#2,#3)} % TODO-nrf rename the commands. Lots of legacy naming around these parts...
+
+\newcommand{\tcheck}[2]{{\sf check}(#1, #2)}
 \renewcommand{\tstr}[1]{{{\sf str}[#1]}}
-\newcommand{\preplace}[3]{{\sf preplace}(#1;#2;#3)}
-\newcommand{\tconcat}[2]{{\sf concat}(#1;#2)} \newcommand{\concat}[2]{{\sf concat}(#1;#2)} % lots of legacy naming around these parts...
+\newcommand{\preplace}[3]{{\sf preplace}(#1,#2,#3)}
+\newcommand{\tconcat}[2]{{\sf concat}(#1,#2)} \newcommand{\concat}[2]{{\sf concat}(#1,#2)} % lots of legacy naming around these parts...
 
@@
 % Meta-theoretic functions
-\newcommand{\lsubst}[3]{{\tt lsubst}(#1;#2;#3)} % This used to render lreplace(...) so there're probably mistkes wherever this command was used now.
-\newcommand{\lreplace}[3]{{\tt lreplace}(#1; #2; #3)}
+\newcommand{\lsubst}[3]{{\tt lsubst}(#1,#2,#3)} % This used to render lreplace(...) so there're probably mistkes wherever this command was used now.
+\newcommand{\lreplace}[3]{{\tt lreplace}(#1, #2, #3)}
 
@@
 Web applications must ultimately command systems like web browsers and database engines using strings.
-Strings constructed using user input that has not been properly sanitized can thus cause command injection vulnerabilities.
+Strings derived from improperly sanitized user input  can thus be a vector for command injection attacks.
 %Structured frameworks for constructing commands centralizes control over string sanitization, but verifying that the framework itself sanitizes user input correctly can still be tricky.
 In this paper, we introduce \emph{regular string types}, which classify strings known statically to be in a specified regular language.
-These types come equipped with common operations like concatenation, substitution and coercion, so they  can be used to implement, in essentially a conventional manner, the parts of a web application or application framework that construct command strings.
+These types come equipped with common operations like concatenation, substitution and coercion, so they  can be used to implement, in essentially a conventional manner, the portions of a web application or web application framework that must directly construct command strings.
 Simple type annotations at key interfaces can be used to statically verify that sanitization has been performed correctly without introducing redundant run-time checks.
@@
 To be practical, adopting a specialized type system like this should not require the adoption of a new programming language.
-Instead, we favor extensible type systems: new type system fragments like this should be distributed as libraries atop a mechanism that guarantees that they can be safely composed.
+Instead, we favor extensible type systems: new type system fragments like this should be implemented as libraries atop a mechanism that guarantees that they can be safely composed.
 We support this by 1) specifying a translation from $\lambda_{RS}$ to a language containing only strings and regular expressions, then, taking Python as such a language, 2) implement the type system together with the translation as a library using \texttt{atlang}, an extensible static type system for Python (being developed by the authors).% represents the most sophisticated application of this library to date.
@@
 Command injection vulnerabilities are among the most common and severe security vulnerabilities in modern web applications \cite{OWASP}.
-They arise because web applications, at their boundaries, control external systems by issuing  commands represented using  strings.
+They arise because web applications, at their boundaries, control external systems using commands represented as  strings.
 For example, web browsers are controlled using HTML and Javascript sent from a server as a string, and database engines execute SQL queries also sent as strings.
-When these commands must include data derived from user input, care must be taken to ensure that the user cannot construct input that subverts the intended command.
+When these commands include data derived from user input, care must be taken to ensure that the user cannot  subverts the intended command by carefully crafting the data they send.
 For example, a  SQL query constructed using string concatenation exposes a SQL injection vulnerability:
@@
 For example, in this case, the developer (or, more often, a framework) might define a function \verb|sanitize| that prepends double quotes (and existing backslashes)  with a backslash, which SQL treats safely.
-Guaranteeing that user input has already been sanitized before it is used to construct a command is challenging.
+Guaranteeing that user input has already been sanitized like this before it is used to construct a command is challenging.
 Note that this function is not idempotent, so it should only be called once.
 
-We observe that most such sanitization techniques can be understood in terms of \emph{regular languages} \cite{cinderella}.
-For example, \verb|name| should be a string in the language described by the regular expression \verb!([^"\]|(\")|(\\))*!
+We observe that many such sanitization techniques can be understood using \emph{regular languages} \cite{cinderella}.
+For example, \verb|name| must be a string in the language described by the regular expression \verb!([^"\]|(\")|(\\))*!
 -- a sequence of characters other than quotation marks and backslashes; these can only appear escaped.
-This concrete pattern syntax can be understood to desugar, in a standard way, to the syntax for regular expressions shown in Figure \ref{fig:regex}, where $r \cdot r$ is sequencing and $r + r$ is  disjunction.
+This concrete syntax for regular expression patterns can be understood to desugar, in a standard way, to the syntax for regular expressions shown in Figure \ref{fig:regex}, where $r \cdot r$ is sequencing and $r + r$ is  disjunction.
 We will work with this ``core'' for simplicity.
@@
 In this paper, we present a static type system that tracks the regular language a string belongs to.
-For example, the output of \verb|sanitize| will be in the regular language described by the regular expression above.
-We write this as $\mathcal{L}\{r\}$.
-We take advantage of closure and decidability properties of regular languages to support a number of useful operations on values of such \emph{regular string types}, including \emph{replacement} of substrings matching a given regular expression.
-These make it possible to implement sanitation protocols like the one just described in an essentially conventional manner.
+For example, the output of \verb|sanitize| will be a string in the regular language described by the regular expression above (we describe such a regular language, following convention, as $\mathcal{L}\{r\}$).
+By leveraging closure and decidability properties of regular languages, the language of a string is tracked through uses of a number of operations, including \emph{replacement} of substrings matching a given regular expression.
+This makes it simple to implement sanitation procedures like the one just described in an essentially conventional manner.
 The result is a system where the fact that a string has been {correctly} sanitized is manifest in its type.
@@
       $\lambda x {:} \sigma . e$ \alt
-      $e~(e)$ \alt
+      $e(e)$ \alt
       $\strin{s}$ \hfill $s \in \Sigma^{*}$ \alt
@@
 \inferrule[S-T-Concat]
-{\sctx \vdash e_1 : \stringin{r_1} \\ \sctx \vdash e_2 : \stringin{r_2}}
-{\sctx \vdash \rsconcat{e_1}{e_2} : \stringin{r_1 \cdot r_2}}
+{\sctx \vdash S_1 : \stringin{r_1} \\ \sctx \vdash S_2 : \stringin{r_2}}
+{\sctx \vdash \rsconcat{S_1}{S_2} : \stringin{r_1 \cdot r_2}}
 
 \inferrule[S-T-Replace]
-{\sctx \vdash e_1 : \stringin{r_1} \\ \sctx \vdash e_2 : \stringin{r_2} \\\\ \lreplace{r}{r_1}{r_2} = r'}
-{\sctx \vdash \sisubst{r}{e_1}{e_2} : \stringin{r'}}
+{\sctx \vdash S_1 : \stringin{r_1} \\ \sctx \vdash S_2 : \stringin{r_2} \\\\ \lreplace{r}{r_1}{r_2} = r'}
+{\sctx \vdash \sisubst{r}{S_1}{S_2} : \stringin{r'}}
 
 \inferrule[S-T-coerce]
-{\sctx \vdash e : \stringin{r'}}
-{\sctx \vdash \coerce{r}{e} : \stringin{r}}
+{\sctx \vdash S : \stringin{r'}}
+{\sctx \vdash \coerce{r}{S} : \stringin{r}}
 
@@
 \inferrule[S-E-Concat]
-{e_1 \sreduces \strin{s_1} \\ e_2 \sreduces \strin{s_2}} 
-{\rsconcat{e_1}{e_2} \sreduces \strin{s_1 s_2}} % ???
+{S_1 \sreduces \strin{s_1} \\ S_2 \sreduces \strin{s_2}} 
+{\rsconcat{S_1}{S_2} \sreduces \strin{s_1 s_2}} % ???
 
 \inferrule[S-E-Replace]
-{e_1 \sreduces \strin{e_1} \\ e_2 \sreduces \strin{s_2} \\ \lsubst{r}{s_1}{s_2} = s} 
-{\sisubst{r}{e_1}{e_2} \sreduces \sistr{s}}
+{S_1 \sreduces \strin{s_1} \\ S_2 \sreduces \strin{s_2} \\ \lsubst{r}{s_1}{s_2} = s} 
+{\sisubst{r}{S_1}{S_2} \sreduces \sistr{s}}
 
@@
 {\coerce{r}{S} \sreduces \strin{s}}
+
+\inferrule[S-E-Coerce-Err]
+{S \sreduces \strin{s} \\ s \not \in \lang{r}}
+{\coerce{r}{S} \err}
 
@@
 \inferrule[Tr-Concat]
-{ \trden{e_1} = \iota_1 \\ \trden{e_2} = \iota_2}
-{\trden{\rsconcat{e_1}{e_2}} = \tconcat{\iota_1}{\iota_2}}
+{ \trden{S_1} = P_1 \\ \trden{S_2} = P_2}
+{\trden{\rsconcat{S_1}{S_2}} = \tconcat{P_1}{P_2}}
 
 \inferrule[Tr-Subst]
-{ \trden{e_1} = \iota_1 \\ \trden{e_2} = \iota_2 }
-{\trden{ \sisubst{r}{e_1}{e_2} } = \tsubst{\rx{r}}{\iota_1}{\iota_2} }
+{ \trden{S_1} = P_1 \\ \trden{S_2} = P_2 }
+{\trden{ \sisubst{r}{S_1}{S_2} } = \tsubst{\rx{r}}{P_1}{P_2} }
 
 \inferrule[Tr-Coerce-Ok]
-{ e : \strin{r} \\ \lang{r'} \subseteq \lang{r}}
-{\trden{ \coerce{r'}{e} } = \tstr{s}}
+{ S : \strin{r} \\ \lang{r'} \subseteq \lang{r}}
+{\trden{ \coerce{r'}{S} } = \tstr{s}}
 
 \inferrule[Tr-Coerce-NotOk]
-{\trden{e} = \iota \\ e : \strin{r} \\ \lang{r'} \not \subseteq \lang{r}}
-{\trden{\coerce{r'}{e}} = \tcheck{\rx{r'}}{\iota}}
+{\trden{S} = P \\ S : \strin{r} \\ \lang{r'} \not \subseteq \lang{r}}
+{\trden{\coerce{r'}{S}} = \tcheck{\rx{r'}}{P}}
 
@@
 \inferrule[P-T-Concat]
-{\tctx \vdash \iota_1 : \str \\ \tctx \vdash \iota_2 : \str}
-{\tctx \vdash \tconcat{\iota_1}{\iota_2} : \str}
+{\tctx \vdash P_1 : \str \\ \tctx \vdash P_2 : \str}
+{\tctx \vdash \tconcat{P_1}{P_2} : \str}
 
 \inferrule[P-T-Replace]
-{\tctx \vdash \iota_1 : \regex \\ \tctx \vdash \iota_2 : \str \\ \tctx \vdash \iota_3 : \str }
-{\tctx \vdash \preplace{\iota_1}{\iota_2}{\iota_3} : \str}
+{\tctx \vdash P_1 : \regex \\ \tctx \vdash P_2 : \str \\ \tctx \vdash P_3 : \str }
+{\tctx \vdash \preplace{P_1}{P_2}{P_3} : \str}
 
 \inferrule[P-T-Check]
-{\tctx \vdash \iota_1 : \regex \\ \tctx \vdash \iota_2 : \str \\ }
-{\tctx \vdash \tcheck{\iota_1}{\iota_2} : \str}
+{\tctx \vdash P_1 : \regex \\ \tctx \vdash P_2 : \str \\ }
+{\tctx \vdash \tcheck{P_1}{P_2} : \str}
 \end{mathpar}
@@
 \inferrule[P-E-Concat]
-{\iota_1 \treduces \tstr{s_1} \\ \iota_2 \treduces \tstr{s_2}} 
-{\tconcat{\iota_1}{\iota_2} \treduces \tstr{s_1 s_2}} % ???
+{P_1 \treduces \tstr{s_1} \\ P_2 \treduces \tstr{s_2}} 
+{\tconcat{P_1}{P_2} \treduces \tstr{s_1 s_2}} % ???
 
 \inferrule[P-E-Replace]
-{\iota_1 \treduces \rx{r} \\ \iota_2 \treduces \tstr{s_2} \\ \iota_3 \treduces \tstr{s_3} \\ \lsubst{r}{s_2}{s_3} = s} 
-{\preplace{\iota_1}{\iota_2}{\iota_3} \treduces \tstr{s}}
+{P_1 \treduces \rx{r} \\ P_2 \treduces \tstr{s_2} \\ P_3 \treduces \tstr{s_3} \\ \lsubst{r}{s_2}{s_3} = s} 
+{\preplace{P_1}{P_2}{P_3} \treduces \tstr{s}}
 
 \inferrule[P-E-Check-OK]
-{\iota_1 \treduces \rx{r} \\ \iota_2 \treduces \strin{s} \\ s \in \lang{r}}
-{\tcheck{\iota_1}{\iota_2} \treduces \tstr{s}}
+{P_1 \treduces \rx{r} \\ P_2 \treduces \strin{s} \\ s \in \lang{r}}
+{\tcheck{P_1}{P_2} \treduces \tstr{s}}
 
 \inferrule[P-E-Check-Err]
-{\iota_1 \treduces \rx{r} \\ \iota_2 \treduces \tstr{s} \\ s \not \in \lang{r}}
-{\tcheck{\iota_1}{\iota_2} \err}
+{P_1 \treduces \rx{r} \\ P_2 \treduces \tstr{s} \\ s \not \in \lang{r}}
+{\tcheck{P_1}{P_2} \err}
 \end{mathpar}
@@
 \begin{lem}
-  If $\sctx \vdash e : \stringin{r}$ then $r$ is a well-formed regular expression.
+  If $\sctx \vdash S : \stringin{r}$ then $r$ is a well-formed regular expression.
 \end{lem}
@@
 \begin{lem}
-If $\tctx \vdash \iota : \regex$ then $\iota \treduces \rx{r}$ such that $r$ is a well-formed regular expression. 
+If $\tctx \vdash P : \regex$ then $P \treduces \rx{r}$ such that $r$ is a well-formed regular expression. 
 \end{lem}
@@
 \begin{thm} 
-  Let $\iota$ be a term in the target language. If $\tctx \vdash \iota : \tau$ 
-  then $\iota \sreduces \iota'$ and $\tctx \vdash \iota' : \tau$.
+  Let $P$ be a term in the target language. If $\tctx \vdash P : \theta$ 
+  then $P \sreduces P'$ and $\tctx \vdash P' : \theta$, or else $P \err$.
 \end{thm}
@@
 \begin{thm}{Type Safety for the String Fragment of $\lambdas$.} \label{thm:scorrect}
-  Let $S$ be a term in the source language. If $\sctx \vdash e : \stringin{r}$ 
-  then $e \sreduces \rstr{s}$ and $\sctx \vdash \rstr{s} : \stringin{r}$.
+  Let $S$ be a term in the source language. If $\sctx \vdash S : \stringin{r}$ 
+  then $S \sreduces \rstr{s}$ and $\sctx \vdash \rstr{s} : \stringin{r}$; or else $S \err$.
 \end{thm}
 \begin{proof}
-By induction on the typing relation.
-The S-T-Concat case requires Lemma \ref{thm:regexprops} 
-and the S-T-Replace case appeals to Lemma \ref{thm:substcorrespondence}.
+By induction on the typing relation, where (a) case holds by lemma \ref{thm:regexprops} in the S-T-Concat case and lemma \ref{thm:substcorrespondence} in the S-T-Replace case..
+The (b) cases hold by unstated, but standard, error propagation rules.
 \end{proof}
@@
 \begin{thm}{Correctness of Input Sanitation for $\lambdas$.}\label{thm:scorrect}
-  If  $\sctx \vdash e : \stringin{r}$ and $e \sreduces \rstr{s}$ then $s \in \lang{r}$.
+  If  $\sctx \vdash S : \stringin{r}$ and $S \sreduces \rstr{s}$ then $s \in \lang{r}$.
 \end{thm}
@@
 \begin{thm}{Translation Correctness.} \label{thm:tcorrect}
-If $\sctx \vdash e : \stringin{r}$ then there exists a $\iota$ such that $\trden{e} = \iota$, 
-$\iota \treduces \tstr{s}$, and $e \sreduces \strin{s}$, or
+If $\sctx \vdash S : \stringin{r}$ then there exists a $P$ such that $\trden{S} = P$ and either: 
+(a) $P \treduces \tstr{s}$ and $S \sreduces \strin{s}$, or
+(b) $P \err$ and $S \err$.
 \end{thm}
@@
 \begin{proof}
-The proof proceeds by induction on the typing relation for $e$ and an appropriate
-choice of $\iota$; in each case, the choice is obvious and the rest of the proof
-proceeds by our type safety theorems as well as an appeal to the induction hypothesis.
+The proof proceeds by induction on the typing relation for $S$ and an appropriate
+choice of $P$; in each case, the choice is obvious. The subcases (a) proceed
+by inversion and appeals to our type safety theorems as well as the induction hypothesis. The subcases (b) proceed
+by the standard error propagation rules omitted for space.
+Throughout the proof, properties from the closure lemma for regular languages
+are necessary.
 \end{proof}
@@
 \begin{thm}{Correctness of Input Sanitation for Translated Terms.}\label{thm:main}
-  If $\trden{e} = \iota$ and $\sctx \vdash e : \stringin{r}$ then $\iota \sreduces \tstr{s}$
+  If $\trden{S} = P$ and $\sctx \vdash S : \stringin{r}$ then either $P \err$ or $P \sreduces \tstr{s}$
   for $s \in \lang{r}$.
@@
 \begin{proof}
-  By theorem \ref{thm:tcorrect}, $\iota \sreduces \tstr{s}$ implies that $e \sreduces \strin{s}$.
-  By theorem \ref{thm:scorrect}, the above property together with the assumption that $e$ is well-typed implies that $s \in \lang{r}$.
+  By theorem \ref{thm:tcorrect}, $P \sreduces \tstr{s}$ implies that $S \sreduces \strin{s}$.
+  By theorem \ref{thm:scorrect}, the above property together with the assumption that $S$ is well-typed implies that $s \in \lang{r}$.
 \end{proof}
@@
 \section{Implementation in {Atlang}}\label{atlang}
+In the previous section, we specified a type system and a translation semantics to a language containing only strings and regular expressions.
+In this section, we take Python to be such a target language.
+
 \begin{figure}\label{fig:example}

### 2014-08-13 sanitation-psp14/sanitation-psp14-short.tex
Cleaned up grammar of target language.

--- 2014-08-13 sanitation-psp14/sanitation-psp14-short.tex
@@
 
+\inferrule[Tr-Abs?]
+{ \sctx \vdash \trden{e} = \iota }
+{\sctx \vdash \trden{ \lambda x.e } = \lambda x . \iota}
+
+\inferrule[Tr-App]
+{\sctx \vdash \trden{e_1} = \iota_1 \\ \sctx \vdash \trden{e_2} = \iota_2}
+{\sctx \vdash \trden{e_1e_2} = \iota_1\iota_2}
+
 \inferrule[Tr-string]
 { \ }
-{\trden{\strin{s}} = \tstr{s}}
+{ \sctx \vdash  \trden{\strin{s}} = \tstr{s}}
 
 \inferrule[Tr-Concat]
-{ \trden{e_1} = \iota_1 \\ \trden{e_2} = \iota_2}
-{\trden{\rsconcat{e_1}{e_2}} = \tconcat{\iota_1}{\iota_2}}
+{ \sctx \vdash \trden{e_1} = \iota_1 \\ \sctx \vdash htrden{e_2} = \iota_2}
+{ \sctx \vdash \trden{\rsconcat{e_1}{e_2}} = \tconcat{\iota_1}{\iota_2} }
 
 \inferrule[Tr-Subst]
-{ \trden{e_1} = \iota_1 \\ \trden{e_2} = \iota_2 }
-{\trden{ \sisubst{r}{e_1}{e_2} } = \tsubst{\rx{r}}{\iota_1}{\iota_2} }
-
-\inferrule[Tr-Coerce-Ok]
-{ e : \strin{r} \\ \lang{r'} \subseteq \lang{r}}
-{\trden{ \coerce{r'}{e} } = \tstr{s}}
-
-\inferrule[Tr-Coerce-NotOk]
-{\trden{e} = \iota \\ e : \strin{r} \\ \lang{r'} \not \subseteq \lang{r}}
-{\trden{\coerce{r'}{e}} = \tcheck{\rx{r'}}{\iota}}
+{ \sctx \vdash  \trden{e_1} = \iota_1 \\ \sctx \vdash  \trden{e_2} = \iota_2 }
+{ \sctx \vdash  \trden{ \sisubst{r}{e_1}{e_2} } = \tsubst{\rx{r}}{\iota_1}{\iota_2} }
+
+\inferrule[Tr-SafeCoerce]
+{ \ }
+{ \sctx \vdash  \trden{ \coerce{r'}{e} } = \trden{e} }
+
+\inferrule[Tr-Check-Ok]
+{ \sctx \vdash e : \stringin{r} \\ \lang{r} \subseteq \lang{r'} } 
+{ \sctx \vdash  \trden{ \rcheck{r'}{e}{x.e_1}{e_2} } = \trden{[e/x]e_1} }
+
+\inferrule[Tr-Check-NotOk]
+{ \sctx \vdash e : \stringin{r} \\ \lang{r} \not \subseteq \lang{r'}}
+{ \sctx \vdash  \trden{ \rcheck{r'}{e}{x.e_1}{e_2} } = \trden{\tcheck{r'}{e_2}} }
 
 \end{mathpar}
-\caption{Translation from source terms (S) to target terms (P).
-The translation is type-directed in the Tr-Coerce cases.}
+\caption{Translation from source terms ($e$) to target terms ($\iota$).
+The translation is type-directed in the Tr-SafeCoerce case.}
 \end{figure*}
@@
 
-<P> ::= $\lambda x . e$ \hfill target terms \alt
-  $e e$ \alt
+<P> ::= $\lambda x . \iota$ \hfill target terms \alt
+  $\iota \iota$ \alt
   $\tstr{s}$ \alt
   $\rx{r}$ \alt
-  $\tconcat{P}{P}$ \alt
-  $\preplace{P}{P}{P}$ \alt
-  $\tcheck{P}{P}$ 
+  $\tconcat{\iota}{\iota}$ \alt
+  $\preplace{\iota}{\iota}{\iota}$ \alt
+  $\tcheck{\iota}{\iota}$ 
   \end{grammar}
-\caption{Syntax for the fragment of our target language, $\lambdap$, containing strings and statically constructed regular expressions.}
+\caption{Syntax for the target language, $\lambdap$, containing strings and statically constructed regular expressions.}
 \label{fig:lcsSyntax}

### 2014-08-13 sanitation-psp14/sanitation-psp14-short.tex
Getting the easy stuff right.
Fixed syntax of lam_rs.

@@
 
-<P> ::= $\lambda x . \iota$ \hfill target terms \alt
-  $\iota \iota$ \alt
+<P> ::= $\lambda x . e$ \hfill target terms \alt
+  $e e$ \alt
   $\tstr{s}$ \alt
   $\rx{r}$ \alt
-  $\tconcat{\iota}{\iota}$ \alt
-  $\preplace{\iota}{\iota}{\iota}$ \alt
-  $\tcheck{\iota}{\iota}$ 
+  $\tconcat{P}{P}$ \alt
+  $\preplace{P}{P}{P}$ \alt
+  $\tcheck{P}{P}$ 
   \end{grammar}
-\caption{Syntax for the target language, $\lambdap$, containing strings and statically constructed regular expressions.}
+\caption{Syntax for the fragment of our target language, $\lambdap$, containing strings and statically constructed regular expressions.}
 \label{fig:lcsSyntax}
@@
 \section{Implementation in {Atlang}}\label{atlang}
-\begin{figure}\label{fig:example}
+In the previous section, we specified a type system and a translation semantics to a language containing only strings and regular expressions.
+In this section, we take Python to be such a target language.
+Python does not have a static type system, however, so to implement these semantics, we will leverage \verb|atlang|, an extensible type system for Python (being developed by the authors).
+By using \verb|atlang|, we can implement these semantics as a library, rather than as a dialect of Python.
+
+\begin{figure}\label{fig:atexample}
 \begin{lstlisting}
+from atlib import fn, string_in
+
 @fn
@@
 \caption{Regular string types in atlang, a library that enables static type checking for Python. }
-\end{figure}\label{fig:impl}
+\end{figure}\label{fig:atimpl}
 \begin{figure}
@@
 
-Figure \ref{fig:example} demosntrates the use of string types in atlang.
+Figure \ref{fig:atexample} demosntrates the use of string types in atlang.
 The sanitize function takes an arbitrary string and returns a string without

### 2014-08-14 sanitation-psp14/sanitation-psp14-short.tex
More editing.
Cleaned up tr rules; various other edits.
Removed counter magic from the tr rule figure
Removed contexts

@@
 
-Waiting for a language designer to build this feature in is unsatisfying.
-Moreover, we would also face a ``chicken-and-egg problem'': justifying its inclusion into a commonly used language requires empirically demonstrating that it is useful, but this is difficult to do if developers have no way to use it in practice.
+Waiting for a language designer to build this feature in is unsatisfying in practice.
+Moreover, we also face a ``chicken-and-egg problem'': justifying its inclusion into a commonly used language benefits from empirical demonstrations that it is useful, but this is difficult to do if developers have no way to use it in practice.
 As such, we take the position that a better path forward for the community is to work within a programming language where such type system fragments can be introduced modularly and orthogonally, as libraries.
@@
       $\sisubst{r}{e}{e}$ \alt
-      $\coerce{r}{e}{x.e}{e}$
+      $\rcoerce{r}{e}$ \alt
+      $\rcheck{r}{e}{x.e}{e}$
 \caption{Syntax of $\lambda_{RS}$.}\label{fig:lamrs}
@@
 \small
-$\fbox{\inferrule{}{\sctx \vdash S : \sigma}}$
+$\fbox{\inferrule{}{\sctx \vdash e : \sigma}}$
 ~~~~$\sctx ::= \emptyset \pipe \sctx, x : \sigma$
 \begin{mathpar}
+\inferrule[S-T-Var]
+{ x:\sigma \in \sctx }
+{ \sctx \vdash s:\sigma}
+
 \inferrule[S-T-Abs]
 {\sctx, x : \sigma_1 \vdash e : \sigma_2}
-{\sctx \vdash \lambda x.e : \sigma_1 \rightarrow \sigma_2}
+{\sctx \vdash \lambda x:\sigma_1.e : \sigma_1 \rightarrow \sigma_2}
 
 \inferrule[S-T-App]
-{\sctx \vdash e_1 : \sigma_1 \\ \sctx \vdash e_2 : \sigma_1 \rightarrow \sigma}
-{\sctx \vdash e_1 e_2 : \sigma}
+{\sctx \vdash e_1 : \sigma_2 \rightarrow \sigma \\ \sctx \vdash e_2 : \sigma_2}
+{\sctx \vdash e_1(e_2) : \sigma}
 
@@
 \small
-$\fbox{\inferrule{}{\tctx \vdash P : \theta}}$
-~~~~$\tctx ::= \emptyset \pipe \tctx, x : \theta$
+$\fbox{\inferrule{}{\tctx \vdash \iota : \tau}}$
+~~~~$\tctx ::= \emptyset \pipe \tctx, x : \tau$
 
 \begin{mathpar}
+\inferrule[P-T-Var]
+{ x:\tau \in \tctx }
+{ \tctx \vdash x:\tau }
+
 \inferrule[P-T-Abs]
 {\tctx, x : \iota_1 \vdash e : \iota_2}
-{\tctx \vdash \lambda x.e : \iota_1 \rightarrow \iota_2}
+{\tctx \vdash \lambda x:\iota_1.e : \iota_1 \rightarrow \iota_2}
 
 \inferrule[P-T-App]
-{\tctx \vdash e_1 : \iota_1 \\ \tctx \vdash e_2 : \iota_1 \rightarrow \iota}
-{\tctx \vdash e_1 e_2 : \iota}
+{\tctx \vdash e_1 : \iota_2 \rightarrow \iota \\ \tctx \vdash e_2 : \iota_2}
+{\tctx \vdash e_1(e_2) : \iota}
 
@@
 Python does not have a static type system, however, so to implement these semantics, we will leverage \verb|atlang|, an extensible type system for Python (being developed by the authors).
-By using \verb|atlang|, we can implement these semantics as a library, rather than as a dialect of Python.
-
-\begin{figure}\label{fig:atexample}
+By using \verb|atlang|, which builds on Python's quotations and reflection facilities, we can implement these semantics as a library, rather than as a new dialect of Python.
+
+\begin{figure}
 \begin{lstlisting}
@@
 \vspace{-10px}
-\caption{Regular string types in atlang, a library that enables static type checking for Python. }
-\end{figure}\label{fig:atimpl}
+\caption{Regular string types in atlang, a library that enables static type checking for Python. }\label{fig:atexample}
+\end{figure}
 \begin{figure}
@@
 \caption{Implementation of the \texttt{string\_in} type constructor in atlang.}
-\end{figure}
+\label{fig:atimpl}\end{figure}
 

### 2014-08-14 sanitation-psp14/sanitation-psp14-short.tex
Remove type annotation from fig 2
Changed check to match tcheck; still not sure how the middle trnaslation of tr-check is supposed to work.

--- 2014-08-13 sanitation-psp14/sanitation-psp14-short.tex
@@
 
-\newcommand{\tcheck}[2]{{\sf check}(#1; #2)}
+\newcommand{\tcheck}[4]{{\sf check}(#1; #2; #3; #4)}
 \renewcommand{\tstr}[1]{{{\sf str}[#1]}}
@@
 % Meta-theoretic functions
-\newcommand{\lsubst}[3]{{\tt lsubst}(#1;#2;#3)} % This used to render lreplace(...) so there're probably mistkes wherever this command was used now.
+\newcommand{\lsubst}[3]{{\tt subst}(#1;#2;#3)} % This used to renderlreplace(...) so there're probably mistkes wherever this command was used now.
 \newcommand{\lreplace}[3]{{\tt lreplace}(#1; #2; #3)}
@@
 % Judgements
-\newcommand{\err}{\ {\sf err}}
 \newcommand{\trden}[1]{\llbracket #1 \rrbracket} % = Translation Denotation.
@@
       $x$ \hfill source terms \alt 
-      $\lambda x {:} \sigma . e$ \alt
+      $\lambda x . e$ \alt
       $e~(e)$ \alt
@@
 \small
-$\fbox{\inferrule{}{\sctx \vdash S : \sigma}}$
+$\fbox{\inferrule{}{\sctx \vdash e : \sigma}}$
 ~~~~$\sctx ::= \emptyset \pipe \sctx, x : \sigma$
 \begin{mathpar}
+\inferrule[S-T-Var]
+{ x:\sigma \in \sctx }
+{ \sctx \vdash x:\sigma}
+
 \inferrule[S-T-Abs]
@@
 \inferrule[S-T-App]
-{\sctx \vdash e_1 : \sigma_1 \\ \sctx \vdash e_2 : \sigma_1 \rightarrow \sigma}
-{\sctx \vdash e_1 e_2 : \sigma}
+{\sctx \vdash e_1 : \sigma_2 \rightarrow \sigma \\ \sctx \vdash e_2 : \sigma_2}
+{\sctx \vdash e_1(e_2) : \sigma}
 
@@
 \small
-$\fbox{\inferrule{}{S \sreduces S}}$
-$\fbox{\inferrule{}{S \err}}$
+$\fbox{\inferrule{}{e \sreduces e}}$
 \begin{mathpar}
@@
 { e_1 \sreduces e_1' \\ e_2 \sreduces e_2' }
-{ e_1 e_2 \sreduces e_1' e_2'}
+{ e_1(e_2) \sreduces e_1'(e_2')}
 
@@
 \end{mathpar}
-\caption{Big step semantics for $\lambdas$.
-Error propagation rules are omitted.}
+\caption{Big step semantics for $\lambdas$}.
 \label{fig:eeval}
@@
 
-\addtocounter{figure}{4}
 \begin{figure*}[t]\label{fig:trans}
 
-$\fbox{\inferrule{}{\trden{S} = P}}$
+$\fbox{\inferrule{}{\sctx \vdash \trden{e} = \iota}}$
 \begin{mathpar}
@@
 {\sctx \vdash \trden{e_1} = \iota_1 \\ \sctx \vdash \trden{e_2} = \iota_2}
-{\sctx \vdash \trden{e_1e_2} = \iota_1\iota_2}
+{\sctx \vdash \trden{e_1(e_2)} = \iota_1(\iota_2)}
 
@@
 \inferrule[Tr-Concat]
-{ \sctx \vdash \trden{e_1} = \iota_1 \\ \sctx \vdash htrden{e_2} = \iota_2}
+{ \sctx \vdash \trden{e_1} = \iota_1 \\ \sctx \vdash \trden{e_2} = \iota_2}
 { \sctx \vdash \trden{\rsconcat{e_1}{e_2}} = \tconcat{\iota_1}{\iota_2} }
@@
 \inferrule[Tr-SafeCoerce]
-{ \ }
-{ \sctx \vdash  \trden{ \coerce{r'}{e} } = \trden{e} }
-
-\inferrule[Tr-Check-Ok]
-{ \sctx \vdash e : \stringin{r} \\ \lang{r} \subseteq \lang{r'} } 
-{ \sctx \vdash  \trden{ \rcheck{r'}{e}{x.e_1}{e_2} } = \trden{[e/x]e_1} }
-
-\inferrule[Tr-Check-NotOk]
-{ \sctx \vdash e : \stringin{r} \\ \lang{r} \not \subseteq \lang{r'}}
-{ \sctx \vdash  \trden{ \rcheck{r'}{e}{x.e_1}{e_2} } = \trden{\tcheck{r'}{e_2}} }
+{ \sctx \vdash \trden{e} = \iota }
+{ \sctx \vdash  \trden{ \coerce{r'}{e} } = \iota }
+
+\inferrule[Tr-Check]
+{ \sctx \vdash \trden{e} = \iota_1 \\ \sctx \vdash \trden{e_1} = \iota_1?? \\ \sctx \vdash \trden{e_2} = \iota_2} 
+{ \sctx \vdash \trden{ \rcheck{r}{e}{x.e_1}{e_2} } = \tcheck{\rx{r}}{\iota}{x.\iota_1}{\iota_2} }
 
@@
 \caption{Translation from source terms ($e$) to target terms ($\iota$).
-The translation is type-directed in the Tr-SafeCoerce case.}
+The translation is type-directed.}
 \end{figure*}
-\addtocounter{figure}{-5}
 
@@
 \small
-$\fbox{\inferrule{}{\tctx \vdash P : \theta}}$
-~~~~$\tctx ::= \emptyset \pipe \tctx, x : \theta$
+$\fbox{\inferrule{}{\tctx \vdash \iota : \tau}}$
+~~~~$\tctx ::= \emptyset \pipe \tctx, x : \tau$
 
 \begin{mathpar}
+\inferrule[P-T-Var]
+{ x:\tau \in \tctx }
+{ \tctx \vdash x:\tau }
+
 \inferrule[P-T-Abs]
-{\tctx, x : \iota_1 \vdash e : \iota_2}
-{\tctx \vdash \lambda x.e : \iota_1 \rightarrow \iota_2}
+{\tctx, x : \tau_1 \vdash \iota_2 : \tau_2}
+{\tctx \vdash \lambda x.\iota_2 : \tau_1 \rightarrow \tau_2}
 
 \inferrule[P-T-App]
-{\tctx \vdash e_1 : \iota_1 \\ \tctx \vdash e_2 : \iota_1 \rightarrow \iota}
-{\tctx \vdash e_1 e_2 : \iota}
+{\tctx \vdash \iota_1 : \tau_2 \rightarrow \tau \\ \tctx \vdash \iota_2 : \tau_2}
+{\tctx \vdash \iota_1(\iota_2) : \iota}
 
@@
 \inferrule[P-T-Check]
-{\tctx \vdash \iota_1 : \regex \\ \tctx \vdash \iota_2 : \str \\ }
-{\tctx \vdash \tcheck{\iota_1}{\iota_2} : \str}
+{\tctx \vdash \iota_1 : \regex \\ \tctx, x:\str \vdash \iota_2 : \sigma \\ \tctx \vdash \iota_3 : \sigma}
+{\tctx \vdash \tcheck{\iota_1}{x.\iota_2}{\iota_3} : \sigma}
 \end{mathpar}
@@
 \small
-$\fbox{\inferrule{}{P \treduces P}}$
-$\fbox{\inferrule{}{P \err}}$
+$\fbox{\inferrule{}{\iota \treduces \iota}}$
 
@@
 \inferrule[P-E-App]
-{ e_1 \sreduces e_1' \\  e_2 \sreduces e_2' }
-{ e_1 e_2 \sreduces e_1' e_2'}
+{ \iota_1 \sreduces \iota_1' \\  \iota_2 \sreduces \iota_2' }
+{ \iota_1(\iota_2) \sreduces \iota_1'(\iota_2')}
 
@@
 \inferrule[P-E-Check-OK]
+{i \treduces \str{s} \\ s \in \lang{r}}
+{\tcheck{r}{i}{x.i_1}{i_2} \treduces [\str{s}/x]i_1}
+
+\inferrule[P-E-Check-NotOK]
+{i \treduces \str{s} \\ s \not \in \lang{r}}
+{\tcheck{r}{i}{x.i_1}{i_2} \treduces i_2}
+
 {\iota_1 \treduces \rx{r} \\ \iota_2 \treduces \strin{s} \\ s \in \lang{r}}
@@
 
-\inferrule[P-E-Check-Err]
-{\iota_1 \treduces \rx{r} \\ \iota_2 \treduces \tstr{s} \\ s \not \in \lang{r}}
-{\tcheck{\iota_1}{\iota_2} \err}
 \end{mathpar}
-\caption{Big step semantics for of $\lambdap$.
-Error propagation rules are omitted.}
+\caption{Big step semantics for of $\lambdap$}.
 \label{fig:ieval}
 \end{figure}
-
-\newpage
-\ 
-\newpage
-\ 
-\newpage
-\
 
@@
 
-<P> ::= $\lambda x . \iota$ \hfill target terms \alt
+<P> ::= $x$  \hfill target terms \alt
+   $\lambda x . \iota$ \alt
    $\iota \iota$ \alt
@@
   $\preplace{\iota}{\iota}{\iota}$ \alt
-  $\tcheck{\iota}{\iota}$ 
+  $\tcheck{\rx{r}}{\iota}{\iota}{\iota}$ 
   \end{grammar}
@@
 
-\begin{defn}[$\tt{lsubst}$]
+\begin{defn}[$\tt{subst}$]
 The relation $\lsubst{r}{s_1}{s_2} = s$ produces a string $s$ in which all substrings of $s_1$ matching $r$ are replaced with $s_2$.
@@
 \begin{proof}
-The theorem follows from the definitions of lsubst and lreplace; note that language substitutions over-approximate string substitutions.
+The theorem follows from the definitions of subst and lreplace; note that language substitutions over-approximate string substitutions.
 \end{proof}

### 2014-08-14 sanitation-psp14/sanitation-psp14-short.tex
Cleanup
Layout

@@
 We will begin in Sec.
-\ref{calculus} by specifying this type system minimally, as a conservative extension of the simply typed lambda calculus called $\lambda_{RC}$.
+\ref{calculus} by specifying this type system minimally, as a conservative extension of the simply typed lambda calculus called $\lambdas$.
 This allows us to specify the guarantees that the type system provides precisely.
@@
 
-% TODO: need to update remainder of intro ...
-To demonstrate this approach, we present a simply typed lambda calculus with \emph{constrained strings}; that is, a set of string types parameterized by regular expressions.
-If $s : \stringin{r}$, then $s$ is a string matching the language $r$.
-Additionally, we include an operation $\rreplace{r}{s_1}{s_2}$ which corresponds to the replace mechanism available in most regular expression libraries; that is, any substring of $s_1$ matching $r$ is replaced with $s_2$.
-The type of this expression is the computed, and is likely ``smaller" or more constrained than the type of $s_1$.
-Libraries, frameworks or functions which construct and execute commands containing input can specify a safe subset $\stringin{r_{\sf spec}}$ of strings, and input sanitation algorithms can construct such a string using ${\sf rreplace}$ or, optionally, by coercion (in which case a runtime check is inserted).
-We also show how this system is translated into a host language containing a regular expression library such that the safety guarantee of the extended language is preserved.
-
-% Summarily, we present a simple type system extension which ensures the absence of input sanitation vulnerabilities by statically checking input sanitation algorithms which use an underlying regular expression library.
-This approach is \emph{composable} in the sense that it is a conservative extension.
-This approach is also \emph{complementary} to existing input sanitation techniques which use string replacement for input sanitation.
-% Our type system for regular string types, $\lambdas$, is defined as an exntension to a simply-typed lambda calculus.
-% Operations on strings are given types according to corresponding, type-level
-% computations on regular expressions. A concatenation operation is defined in 
-% order to demonstrate this principle, but the premier operation of our system
-% is \emph{replacement}, which captures a rich set of input sanitation techniques
-% including escaped characters and encodings. In addition to defining this system,
-% we present a correctness-preserving compilation of $\lambdas$ into an underlying
-% language with ordinary strings as a regular expression library.
-
-% In addition to this theoretical contribution, we also explain how our extension
-% may be implemented in the Ace Programming Language, which is itself an extension
-% of Python -- a popular web programming language.
-
-% Summarily, we present a simple type system extension which ensures the absence
-% of input sanitation vulnerabilities by statically checking the correctness and 
-% correct use of input sanitation algorithms. This approach is \emph{composable}
-% in the sense that it is a conservative extension. This approach is \emph{complementary}
-% to existing input sanitation techniques that use regular expressions for input sanitation.
-
-%%
-% This is the paragraph which ties the paper back to the CFP.
-%
-% For refernce, here's a copy of the cfp:
-%The importance of security and privacy in software engineering is now a pressing concern. 
-%Over the last decade, many programmers have recognized the importance of including 
-%security and privacy requirements at project start. This has led to efforts 
-%in secure/security engineering and privacy engineering, which focus on guidelines and 
-%best practices that can be used at the design stage to create safer code. 
-%Unfortunately, these disciplines are not pervasive and still in their infancy.
-%The goal of this workshop is to make security and privacy first class citizens in programming and programming languages.
-
-%This forum aims to gather research and industry professionals to further the discussion on how to:
-%   codify the principles from secure and privacy engineering into programming language constructs and or tools, 
-%   create programming languages that have security and privacy as foundational tenets, and 
-%   create/codify constructs or tools that enables secure and privacy-preserving (business) operations.
-%   All submissions connected to the goals above, and their associated topics, are encouraged.
-%%
-
-%\subsection{Outline}
-%
-%An outline of this paper follows:
-%
-%\begin{itemize}
-%  \item In \S 2, we define the type system which is embedded in Ace. We include a type safety proof for the string
-%    segment of this language and prove the correctness of a translation to an underlying langguage $P$. In our theory,
-%    $P$ is a simply typed lambda calculus equipped with a minimal regular expression library; in an implementation, $P$
-%    stands in for Python or another underlying general-purpose programming language.
-%  \item In \S 3, we discuss our implemention of this translation as a type system extension  within the Ace programming language.
-%\end{itemize}
-
 \section{Regular String Types, Minimally}\label{calculus}
@@
 
-\begin{figure}[t]
+\newcommand{\sctx}{\Psi} % Context for external typing
+\newcommand{\tctx}{\Theta} % Context for internal typing
+
+\newcommand{\ereduces}{\Downarrow}
+
+\begin{figure}[h]
 \begin{grammar}
@@
 
-\begin{figure}[t]
+\begin{figure}[h]
 \begin{grammar}
@@
 
-\newcommand{\sctx}{\Psi} % Context for external typing
-\newcommand{\tctx}{\Theta} % Context for internal typing
-
-\newcommand{\ereduces}{\Downarrow}
-
-\begin{figure}[t]\label{fig:lambdas}
-\small
-$\fbox{\inferrule{}{\sctx \vdash e : \sigma}}$
-~~~~$\sctx ::= \emptyset \pipe \sctx, x : \sigma$
-\begin{mathpar}
-\inferrule[S-T-Var]
-{ x:\sigma \in \sctx }
-{ \sctx \vdash x:\sigma}
-
-\inferrule[S-T-Abs]
-{\sctx, x : \sigma_1 \vdash e : \sigma_2}
-{\sctx \vdash \lambda x.e : \sigma_1 \rightarrow \sigma_2}
-
-\inferrule[S-T-App]
-{\sctx \vdash e_1 : \sigma_2 \rightarrow \sigma \\ \sctx \vdash e_2 : \sigma_2}
-{\sctx \vdash e_1(e_2) : \sigma}
-
-\inferrule[S-T-Stringin-I]
-{s \in \lang{r}}
-{\sctx \vdash \strin{s} : \stringin{r}}
-
-\inferrule[S-T-Concat]
-{\sctx \vdash e_1 : \stringin{r_1} \\ \sctx \vdash e_2 : \stringin{r_2}}
-{\sctx \vdash \rsconcat{e_1}{e_2} : \stringin{r_1 \cdot r_2}}
-
-\inferrule[S-T-Replace]
-{\sctx \vdash e_1 : \stringin{r_1} \\ \sctx \vdash e_2 : \stringin{r_2} \\\\ \lreplace{r}{r_1}{r_2} = r'}
-{\sctx \vdash \sisubst{r}{e_1}{e_2} : \stringin{r'}}
-
-\inferrule[S-T-SafeCoerce]
-{\sctx \vdash e : \stringin{r'} \\ \lang{r'} \subseteq \lang{r}}
-{\sctx \vdash \coerce{r}{e} : \stringin{r}}
-
-\inferrule[S-T-Check]
-{\sctx \vdash e_0 : \stringin{r_0} \\ \sctx, x:\stringin{r} \vdash e_1 : \sigma \\ \sctx \vdash e_2 : \sigma}
-{\sctx \vdash \rcheck{r}{e_0}{x.e_1}{e_2} : \sigma}
-
-\end{mathpar}
-
-\caption{Typing rules for $\lambdas$. 
-The typing context $\sctx$ is standard.}
-\label{fig:etyping}
-\end{figure}
-
-\begin{figure}[t]
-\small
-$\fbox{\inferrule{}{e \sreduces e}}$
-\begin{mathpar}
-\inferrule[S-E-Abs]                                                             
-{ \ }                                                                           
-{\lambda x.e \sreduces \lambda x.e}
-
-\inferrule[S-E-App]
-{ e_1 \sreduces e_1' \\ e_2 \sreduces e_2' }
-{ e_1(e_2) \sreduces e_1'(e_2')}
-
-\inferrule[S-E-RStr]
-{ \ }
-{\strin{s} \sreduces \strin{s}}
-
-\inferrule[S-E-Concat]
-{e_1 \sreduces \strin{s_1} \\ e_2 \sreduces \strin{s_2}} 
-{\rsconcat{e_1}{e_2} \sreduces \strin{s_1 s_2}} % ???
-
-\inferrule[S-E-Replace]
-{e_1 \sreduces \strin{e_1} \\ e_2 \sreduces \strin{s_2} \\ \lsubst{r}{s_1}{s_2} = s} 
-{\sisubst{r}{e_1}{e_2} \sreduces \sistr{s}}
-
-\inferrule[S-E-SafeCoerce]
-{e \sreduces \strin{s}}
-{\coerce{r}{e} \sreduces \strin{s}}
-
-\inferrule[S-E-Check-Ok]
-{e \sreduces \strin{s} \\ s \in \lang{r}}
-{
-  \rcheck{r}{e}{x.e_1}{e_2} \sreduces [\strin{s} / x]e_1
-}
-
-\inferrule[S-E-Check-NotOk]
-{
-  e \sreduces \strin{s} \\ s \not \in \lang{r}
-}
-{
-  \rcheck{r}{e}{x.e_1}{e_2} \sreduces e_2
-}
-
-\end{mathpar}
-\caption{Big step semantics for $\lambdas$}.
-\label{fig:eeval}
-\end{figure}
-
 In this section, we define a minimal typed lambda calculus with regular string types called $\lambda_{RS}$.
@@
 \begin{itemize}
-  \item Page 3 contains a definition of $\lambdas, \lambdap$ and the translation from $\lambdas$ to $\lambdap$. Grammar follows immediately at the top of page 4.
-  \item In \S 2.1 we state some properties about regular expressions which are needed in our correctness proofs.
-   \item In \S 2.2 we prove type safety for $\lambdap$ as well as both type safety and correctness for $\lambdas$.
-  \item In \S 2.3 we prove that translation preserves the correctness result about $\lambdas$.
+  \item In \S 2.1-2.3, we define $\lambdas$, $\lambdap$ and the translation from one to the other.
+  \item In \S 2.4 we state some properties about regular expressions which are needed in our correctness proofs.
+   \item In \S 2.5 we prove type safety for $\lambdap$ as well as both type safety and correctness for $\lambdas$.
+  \item In \S 2.6 we prove that translation preserves the correctness result about $\lambdas$.
 \end{itemize}
 
-%The $\lambdas$ language is characterized by a type of strings indexed by regular
-%expressions, together with operations on such strings which correspond to common
-%input sanitation patterns.
-%This section presents the grammar, typing rules and operational semantics for
-%$\lambdas$ as well as an underlying language $\lambdap$.
-%
-%The system $\lambdas$ is the simply typed lambda calculus extended with \emph{regular expression types}, which are string
-%types ensuring a string belongs to a specified language. For instance, 
-%$S : \stringin{r}$ reads ``$s$ is a string matching $r$".
-%the system includes an operation for replacing all instances of a pattern $r$ in a string $s_1$ with another string $s_2$. 
-%Input sanitation algorithms -- as implemented by developers or within popular libraries and frameworks -- are often implemented in terms of this replace operation.
-%
-%The language $\lambdap$ is a simple functional language extended with a minimal regular expression library.
-%Any general purpose programming language could stand in for $\lambdap$; for instance, SML or Python.
-%In an implementation, our correctness results are modulo the underlying language's correct implementation of regular expression matching (see P-E-Replace).
-%
-%Finally, we define a translation from our type system $\lambdas$ into $\lambdap$ which preserves the safety guarnatee codified in the string types of $\lambdas$.
-Because our extension is conservative and its guarnatees are preserved under translation to an underlying language, it is highly composable with other analyses.
-%
-%Unfortunately, we are unable to present full proofs in this paper due to space constraints; however, proofs are given in full in the accompanying technical report?.
-%%Note that we never insert an internal check where the type of the string implies
-%%that a check must succeed. Furthermore, expensive calculations (such as language
-%%inclusion or ${\sf rreplace}$) all occur at compile time.
-%%Since sanitation problems are generally expressible with smalle, simple expressions,
-%%we believe that the compile-time overhead is not significant enough to prohibiour target usecase. Informal experimentation with our implementation seems to
-%%support this assertion.
-
-% Begin Grammar 
-% Removes the < > from grammar productions.
-
-% \begin{figure}[h]
-% \begin{grammar}
-
-% <$\theta$> ::= $\theta \rightarrow \theta$ \hfill target types \alt
-% $\str$ \alt $\regex$
-
-% <P> ::= $\lambda x . e$ \hfill target terms \alt
-%   $e e$ \alt
-%   $\tstr{s}$ \alt
-%   $\rx{r}$ \alt
-%   $\tconcat{P}{P}$ \alt
-%   $\preplace{P}{P}{P}$ \alt
-%   $\tcheck{P}{P}$ 
-%   \end{grammar}
-% \caption{Syntax for the fragment of our target language, $\lambdap$, containing strings and statically constructed regular expressions.}
-% \label{fig:lcsSyntax}
-% \end{figure}
-
-%The $\lcs$ language gives static semantics for common regular expression library
-%functions. In this treatment, we include concatenation and filtering.
-%The ${\tt filter}$ function removes all instances of a regular expression in a string,
-%while concatenation ($+$) concatenates two strings.
-%
-%\subsection{Typing}
-%
-%The $\strin[r]$ type is parameterized by regular expressions; if $e:\strin[r]$,
-%then $e \in r$. Mapping from an arbitrary $\str$ to a $\strin[r]$ requires
-%defining an algorithm -- in terms of filter -- for converting a $\strin[.*]$
-%into a $\strin[r]$. The static semantics of the language defines the types of
-%operations on regular expressions in terms of well-understood properties about
-%regular lanugages; we recall these properties in section 3.
-%
-%\subsection{Dynamics}
-%
-%There are two evaluation judgements: $e:T \reduces e'$ and $e:T \treduces i$.
-%The $\reduces$ relation is between $\lcs$ expressions, while the $\treduces$
-%relation is a mapping from $\lcs$ expressions into internal language expressions
-%$i$ such that $i \ival$.
-%
-%Safety of the evaluation relation depends upon an injective mapping from $\lcs$ types info
-%internal language types. This relation, $h$, is defined below.
-%
-%\subsection{Type Safety}
-%
-%The type safety proof relies upon some assumptions about the type system and
-%dynamics of the internal language, as well as some properties of regular
-%languages.
-%
-%There must exist a translation from $\lcs$ types to the types of the internal
-%language. For the remainder of this paper, we call the type translation function
-%$h$.
-%
-%\begin{defn}[Type Translation Function $h$]
-%  The type translation function $h : Type \rightarrow IType$ is defined as follows:
-%  \begin{itemize}
-%    \item $\forall r. h(\strin[r]) = istr$
-%    \item $h(\str) = istr$
-%  \end{itemize}
-%\end{defn}
-%
-%Additionally, we assume that the internal language contains an implementation of strings,
-%together with operations for concatenation and filtering by regular expression.
-%
-%\begin{defn}[Types of internal values]
-%  Let $\tstrf{s}$ range over string literals and $r$ over regular expressions.
-%  Internal values are typed as follows:
-%\begin{itemize}
-%  \item If $e=\tstrf{s}$ then $e:istr$.
-%  \item If $e=\ifilter{r}{\tstrf{s}}$ then $e:istr$.
-%  \item If $e=\tstrf{s_1}+\tstrf{s_2}$ then $e:istr$.
-%\end{itemize}
-%\end{defn}
-%
-%For simplicity, we assume a fixed translation from
-%$\lcs$ regular expressions to regular expressions recognizable by the internal
-%language's regex library (in practice, a fixed translation is acceptable.)
-%To summarize, we assume an internal language containing 
-%a string type together with operations for string concatentation and filtering. We
-%expect closure over strings for both operations.
-%Finally, recall that ${\tt ifilter}$ is only needed for dynamic casts, which may be removed without
-%descreasing the expressivity or even usability of the language.
-%Finally, the semantics of the filter function are defined in terms of ${\tt rl\_filter}$,
-%which is a static version of ${\tt ifilter}$.
-%
-
-% I can't achieve the correct latout here. We should have something like:
-%       Lam_S Synax         Lam_P Syntax
-%       Lam_S Sematics      LAm_P Semantics
-%                Translation rules
-% Or, alternatively, we can flip reflect things so that the entire page
-% reads top-down (Lam_S syntax Lam_S smantics horizontally aligned w/ one
-% another).
-
-% Unfortunately, I can't get that to work. So, we're settling for something
-% slightly ugly, where the fitures appear in the incorrect order and the 
-% translation rules are on top.
-
-% The translations rules *just won't* go to the bottom of the page, so 
-% instead I am placing them at the top. Since the figures are out of order,
-% it's even more important that we get the number gith. therefore, the counter
-% is modified before and after this figure.
-
-\begin{figure*}[t]\label{fig:trans}
-
-$\fbox{\inferrule{}{\sctx \vdash \trden{e} = \iota}}$
-\begin{mathpar}
-
-\inferrule[Tr-Abs?]
-{ \sctx \vdash \trden{e} = \iota }
-{\sctx \vdash \trden{ \lambda x.e } = \lambda x . \iota}
-
-\inferrule[Tr-App]
-{\sctx \vdash \trden{e_1} = \iota_1 \\ \sctx \vdash \trden{e_2} = \iota_2}
-{\sctx \vdash \trden{e_1(e_2)} = \iota_1(\iota_2)}
-
-\inferrule[Tr-string]
-{ \ }
-{ \sctx \vdash  \trden{\strin{s}} = \tstr{s}}
-
-\inferrule[Tr-Concat]
-{ \sctx \vdash \trden{e_1} = \iota_1 \\ \sctx \vdash \trden{e_2} = \iota_2}
-{ \sctx \vdash \trden{\rsconcat{e_1}{e_2}} = \tconcat{\iota_1}{\iota_2} }
-
-\inferrule[Tr-Subst]
-{ \sctx \vdash  \trden{e_1} = \iota_1 \\ \sctx \vdash  \trden{e_2} = \iota_2 }
-{ \sctx \vdash  \trden{ \sisubst{r}{e_1}{e_2} } = \tsubst{\rx{r}}{\iota_1}{\iota_2} }
-
-\inferrule[Tr-SafeCoerce]
-{ \sctx \vdash \trden{e} = \iota }
-{ \sctx \vdash  \trden{ \coerce{r'}{e} } = \iota }
-
-\inferrule[Tr-Check]
-{ \sctx \vdash \trden{e} = \iota_1 \\ \sctx \vdash \trden{e_1} = \iota_1?? \\ \sctx \vdash \trden{e_2} = \iota_2} 
-{ \sctx \vdash \trden{ \rcheck{r}{e}{x.e_1}{e_2} } = \tcheck{\rx{r}}{\iota}{x.\iota_1}{\iota_2} }
-
-\end{mathpar}
-\caption{Translation from source terms ($e$) to target terms ($\iota$).
-The translation is type-directed.}
-\end{figure*}
-
 %
@@
 %
-
-\newpage
-
-\begin{figure}[t]\label{fig:lambdap}
-\small
-$\fbox{\inferrule{}{\tctx \vdash \iota : \tau}}$
-~~~~$\tctx ::= \emptyset \pipe \tctx, x : \tau$
-
-\begin{mathpar}
-\inferrule[P-T-Var]
-{ x:\tau \in \tctx }
-{ \tctx \vdash x:\tau }
-
-\inferrule[P-T-Abs]
-{\tctx, x : \tau_1 \vdash \iota_2 : \tau_2}
-{\tctx \vdash \lambda x.\iota_2 : \tau_1 \rightarrow \tau_2}
-
-\inferrule[P-T-App]
-{\tctx \vdash \iota_1 : \tau_2 \rightarrow \tau \\ \tctx \vdash \iota_2 : \tau_2}
-{\tctx \vdash \iota_1(\iota_2) : \iota}
-
-\inferrule[P-T-String]
-{ \ }
-{\tctx \vdash \tstr{s} : \str}
-
-\inferrule[P-T-Regex]
-{ \ }
-{\tctx \vdash \rx{r} : \regex}
-
-\inferrule[P-T-Concat]
-{\tctx \vdash \iota_1 : \str \\ \tctx \vdash \iota_2 : \str}
-{\tctx \vdash \tconcat{\iota_1}{\iota_2} : \str}
-
-\inferrule[P-T-Replace]
-{\tctx \vdash \iota_1 : \regex \\ \tctx \vdash \iota_2 : \str \\ \tctx \vdash \iota_3 : \str }
-{\tctx \vdash \preplace{\iota_1}{\iota_2}{\iota_3} : \str}
-
-\inferrule[P-T-Check]
-{\tctx \vdash \iota_1 : \regex \\ \tctx, x:\str \vdash \iota_2 : \sigma \\ \tctx \vdash \iota_3 : \sigma}
-{\tctx \vdash \tcheck{\iota_1}{x.\iota_2}{\iota_3} : \sigma}
-\end{mathpar}
-\caption{Typing rules for $\lambdap$.
-The typing context $\tctx$ is standard.}
-\label{fig:ityping}
-\end{figure}
-
-\begin{figure}[t]
-\small
-$\fbox{\inferrule{}{\iota \treduces \iota}}$
-
-\begin{mathpar}
-\inferrule[P-E-Abs]
-{ \ }
-{\lambda x.e \sreduces \lambda x.e}
-
-\inferrule[P-E-App]
-{ \iota_1 \sreduces \iota_1' \\  \iota_2 \sreduces \iota_2' }
-{ \iota_1(\iota_2) \sreduces \iota_1'(\iota_2')}
-
-  \inferrule[P-E-Str]
-{ \ }
-{\tstr{s} \treduces \tstr{s}}
-
-\inferrule[P-E-Rx]
-{ \ }
-{\rx{r} \treduces \rx{r}}
-
-\inferrule[P-E-Concat]
-{\iota_1 \treduces \tstr{s_1} \\ \iota_2 \treduces \tstr{s_2}} 
-{\tconcat{\iota_1}{\iota_2} \treduces \tstr{s_1 s_2}} % ???
-
-\inferrule[P-E-Replace]
-{\iota_1 \treduces \rx{r} \\ \iota_2 \treduces \tstr{s_2} \\ \iota_3 \treduces \tstr{s_3} \\ \lsubst{r}{s_2}{s_3} = s} 
-{\preplace{\iota_1}{\iota_2}{\iota_3} \treduces \tstr{s}}
-
-\inferrule[P-E-Check-OK]
-{i \treduces \str{s} \\ s \in \lang{r}}
-{\tcheck{r}{i}{x.i_1}{i_2} \treduces [\str{s}/x]i_1}
-
-\inferrule[P-E-Check-NotOK]
-{i \treduces \str{s} \\ s \not \in \lang{r}}
-{\tcheck{r}{i}{x.i_1}{i_2} \treduces i_2}
-
-{\iota_1 \treduces \rx{r} \\ \iota_2 \treduces \strin{s} \\ s \in \lang{r}}
-{\tcheck{\iota_1}{\iota_2} \treduces \tstr{s}}
-
-\end{mathpar}
-\caption{Big step semantics for of $\lambdap$}.
-\label{fig:ieval}
-\end{figure}
 
@@
 % \end{figure}
-
-\begin{figure}[h]
-\begin{grammar}
-
-<$\theta$> ::= $\theta \rightarrow \theta$ \hfill target types \alt
-$\str$ \alt $\regex$
-
-<P> ::= $x$  \hfill target terms \alt
-   $\lambda x . \iota$ \alt
-   $\iota \iota$ \alt
-  $\tstr{s}$ \alt
-  $\rx{r}$ \alt
-  $\tconcat{\iota}{\iota}$ \alt
-  $\preplace{\iota}{\iota}{\iota}$ \alt
-  $\tcheck{\rx{r}}{\iota}{\iota}{\iota}$ 
-  \end{grammar}
-\caption{Syntax for the target language, $\lambdap$, containing strings and statically constructed regular expressions.}
-\label{fig:lcsSyntax}
-\end{figure}
 
@@
     expoiting the decidability of language inclusion.
+  \item The check rules allow casts between strings that cannot be guaranteed at compile time, but might pass a runtime check.
 \end{itemize}
 
+\newpage
+
+\begin{figure}[t]\label{fig:lambdap}
+\small
+$\fbox{\inferrule{}{\tctx \vdash \iota : \tau}}$
+~~~~$\tctx ::= \emptyset \pipe \tctx, x : \tau$
+
+\begin{mathpar}
+\inferrule[P-T-Var]
+{ x:\tau \in \tctx }
+{ \tctx \vdash x:\tau }
+
+\inferrule[P-T-Abs]
+{\tctx, x : \tau_1 \vdash \iota_2 : \tau_2}
+{\tctx \vdash \lambda x.\iota_2 : \tau_1 \rightarrow \tau_2}
+
+\inferrule[P-T-App]
+{\tctx \vdash \iota_1 : \tau_2 \rightarrow \tau \\ \tctx \vdash \iota_2 : \tau_2}
+{\tctx \vdash \iota_1(\iota_2) : \iota}
+
+\inferrule[P-T-String]
+{ \ }
+{\tctx \vdash \tstr{s} : \str}
+
+\inferrule[P-T-Regex]
+{ \ }
+{\tctx \vdash \rx{r} : \regex}
+
+\inferrule[P-T-Concat]
+{\tctx \vdash \iota_1 : \str \\ \tctx \vdash \iota_2 : \str}
+{\tctx \vdash \tconcat{\iota_1}{\iota_2} : \str}
+
+\inferrule[P-T-Replace]
+{\tctx \vdash \iota_1 : \regex \\ \tctx \vdash \iota_2 : \str \\ \tctx \vdash \iota_3 : \str }
+{\tctx \vdash \preplace{\iota_1}{\iota_2}{\iota_3} : \str}
+
+\inferrule[P-T-Check]
+{\tctx \vdash \iota_r : \regex \\ \tctx \vdash \iota_1 : \str \\ \tctx, x:\str \vdash \iota_2 : \sigma \\ \tctx \vdash \iota_3 : \sigma}
+{\tctx \vdash \tcheck{\iota_r}{\iota_1}{x . \iota_2}{\iota_3} : \sigma}
+\end{mathpar}
+\caption{Typing rules for $\lambdap$.
+The typing context $\tctx$ is standard.}
+\label{fig:ityping}
+\end{figure}
+
+\begin{figure}[t]
+\small
+$\fbox{\inferrule{}{\iota \treduces \iota}}$
+
+\begin{mathpar}
+\inferrule[P-E-Abs]
+{ \ }
+{\lambda x.e \sreduces \lambda x.e}
+
+\inferrule[P-E-App]
+{ \iota_1 \sreduces \iota_1' \\  \iota_2 \sreduces \iota_2' }
+{ \iota_1(\iota_2) \sreduces \iota_1'(\iota_2')}
+
+  \inferrule[P-E-Str]
+{ \ }
+{\tstr{s} \treduces \tstr{s}}
+
+\inferrule[P-E-Rx]
+{ \ }
+{\rx{r} \treduces \rx{r}}
+
+\inferrule[P-E-Concat]
+{\iota_1 \treduces \tstr{s_1} \\ \iota_2 \treduces \tstr{s_2}} 
+{\tconcat{\iota_1}{\iota_2} \treduces \tstr{s_1 s_2}} % ???
+
+\inferrule[P-E-Replace]
+{\iota_1 \treduces \rx{r} \\ \iota_2 \treduces \tstr{s_2} \\ \iota_3 \treduces \tstr{s_3} \\ \lsubst{r}{s_2}{s_3} = s} 
+{\preplace{\iota_1}{\iota_2}{\iota_3} \treduces \tstr{s}}
+
+\inferrule[P-E-Check-OK]
+{\iota \treduces \tstr{s} \\ s \in \lang{r}}
+{\tcheck{\rx{r}}{\iota}{x.\iota_1}{\iota_2} \treduces [\tstr{s}/x]\iota_1}
+
+\inferrule[P-E-Check-NotOK]
+{\iota \treduces \tstr{s} \\ s \not \in \lang{r}}
+{\tcheck{\rx{r}}{\iota}{x.\iota_1}{\iota_2} \treduces \iota_2}
+\end{mathpar}
+\caption{Big step semantics for of $\lambdap$}.
+\label{fig:ieval}
+\end{figure}
+
+\begin{figure}\label{fig:lambdas}
+\small
+$\fbox{\inferrule{}{\sctx \vdash e : \sigma}}$
+~~~~$\sctx ::= \emptyset \pipe \sctx, x : \sigma$
+\begin{mathpar}
+\inferrule[S-T-Var]
+{ x:\sigma \in \sctx }
+{ \sctx \vdash x:\sigma}
+
+\inferrule[S-T-Abs]
+{\sctx, x : \sigma_1 \vdash e : \sigma_2}
+{\sctx \vdash \lambda x.e : \sigma_1 \rightarrow \sigma_2}
+
+\inferrule[S-T-App]
+{\sctx \vdash e_1 : \sigma_2 \rightarrow \sigma \\ \sctx \vdash e_2 : \sigma_2}
+{\sctx \vdash e_1(e_2) : \sigma}
+
+\inferrule[S-T-Stringin-I]
+{s \in \lang{r}}
+{\sctx \vdash \strin{s} : \stringin{r}}
+
+\inferrule[S-T-Concat]
+{\sctx \vdash e_1 : \stringin{r_1} \\ \sctx \vdash e_2 : \stringin{r_2}}
+{\sctx \vdash \rsconcat{e_1}{e_2} : \stringin{r_1 \cdot r_2}}
+
+\inferrule[S-T-Replace]
+{\sctx \vdash e_1 : \stringin{r_1} \\ \sctx \vdash e_2 : \stringin{r_2} \\\\ \lreplace{r}{r_1}{r_2} = r'}
+{\sctx \vdash \sisubst{r}{e_1}{e_2} : \stringin{r'}}
+
+\inferrule[S-T-SafeCoerce]
+{\sctx \vdash e : \stringin{r'} \\ \lang{r'} \subseteq \lang{r}}
+{\sctx \vdash \coerce{r}{e} : \stringin{r}}
+
+\inferrule[S-T-Check]
+{\sctx \vdash e_0 : \stringin{r_0} \\ \sctx, x:\stringin{r} \vdash e_1 : \sigma \\ \sctx \vdash e_2 : \sigma}
+{\sctx \vdash \rcheck{r}{e_0}{x.e_1}{e_2} : \sigma}
+
+\end{mathpar}
+
+\caption{Typing rules for $\lambdas$. 
+The typing context $\sctx$ is standard.}
+\label{fig:etyping}
+\end{figure}
+
+\begin{figure}
+\small
+$\fbox{\inferrule{}{e \sreduces e}}$
+\begin{mathpar}
+\inferrule[S-E-Abs]                                                             
+{ \ }                                                                           
+{\lambda x.e \sreduces \lambda x.e}
+
+\inferrule[S-E-App]
+{ e_1 \sreduces e_1' \\ e_2 \sreduces e_2' }
+{ e_1(e_2) \sreduces e_1'(e_2')}
+
+\inferrule[S-E-RStr]
+{ \ }
+{\strin{s} \sreduces \strin{s}}
+
+\inferrule[S-E-Concat]
+{e_1 \sreduces \strin{s_1} \\ e_2 \sreduces \strin{s_2}} 
+{\rsconcat{e_1}{e_2} \sreduces \strin{s_1 s_2}} % ???
+
+\inferrule[S-E-Replace]
+{e_1 \sreduces \strin{e_1} \\ e_2 \sreduces \strin{s_2} \\ \lsubst{r}{s_1}{s_2} = s} 
+{\sisubst{r}{e_1}{e_2} \sreduces \sistr{s}}
+
+\inferrule[S-E-SafeCoerce]
+{e \sreduces \strin{s}}
+{\coerce{r}{e} \sreduces \strin{s}}
+
+\inferrule[S-E-Check-Ok]
+{e \sreduces \strin{s} \\ s \in \lang{r}}
+{
+  \rcheck{r}{e}{x.e_1}{e_2} \sreduces [\strin{s} / x]e_1
+}
+
+\inferrule[S-E-Check-NotOk]
+{
+  e \sreduces \strin{s} \\ s \not \in \lang{r}
+}
+{
+  \rcheck{r}{e}{x.e_1}{e_2} \sreduces e_2
+}
+
+\end{mathpar}
+\caption{Big step semantics for $\lambdas$}.
+\label{fig:eeval}
+\end{figure}
+
+\newpage
+\ 
+\newpage
+\ 
+\newpage
+
 \subsection{Definition of $\lambdap$}
+
+\begin{figure}[h]
+\begin{grammar}
+
+<$\theta$> ::= $\theta \rightarrow \theta$ \hfill target types \alt
+$\str$ \alt $\regex$
+
+<P> ::= $x$  \hfill target terms \alt
+   $\lambda x . \iota$ \alt
+   $\iota \iota$ \alt
+  $\tstr{s}$ \alt
+  $\rx{r}$ \alt
+  $\tconcat{\iota}{\iota}$ \alt
+  $\preplace{\iota}{\iota}{\iota}$ \alt
+  $\tcheck{\rx{r}}{\iota}{\iota}{\iota}$ 
+  \end{grammar}
+\caption{Syntax for the target language, $\lambdap$, containing strings and statically constructed regular expressions.}
+\label{fig:lcsSyntax}
+\end{figure}
 
@@
 \subsection{Definition of Translation}
+
+\begin{figure}[h]\label{fig:trans}
+
+$\fbox{\inferrule{}{\sctx \vdash \trden{e} = \iota}}$
+\begin{mathpar}
+
+\inferrule[Tr-Abs]
+{ \sctx \vdash \trden{e} = \iota }
+{\sctx \vdash \trden{ \lambda x.e } = \lambda x . \iota}
+
+\inferrule[Tr-App]
+{\sctx \vdash \trden{e_1} = \iota_1 \\ \sctx \vdash \trden{e_2} = \iota_2}
+{\sctx \vdash \trden{e_1(e_2)} = \iota_1(\iota_2)}
+
+\inferrule[Tr-string]
+{ \ }
+{ \sctx \vdash  \trden{\strin{s}} = \tstr{s}}
+
+\inferrule[Tr-Concat]
+{ \sctx \vdash \trden{e_1} = \iota_1 \\ \sctx \vdash \trden{e_2} = \iota_2}
+{ \sctx \vdash \trden{\rsconcat{e_1}{e_2}} = \tconcat{\iota_1}{\iota_2} }
+
+\inferrule[Tr-Subst]
+{ \sctx \vdash  \trden{e_1} = \iota_1 \\ \sctx \vdash  \trden{e_2} = \iota_2 }
+{ \sctx \vdash  \trden{ \sisubst{r}{e_1}{e_2} } = \tsubst{\rx{r}}{\iota_1}{\iota_2} }
+
+\inferrule[Tr-SafeCoerce]
+{ \sctx \vdash \trden{e} = \iota }
+{ \sctx \vdash  \trden{ \coerce{r'}{e} } = \iota }
+
+\inferrule[Tr-Check]
+{ \sctx \vdash \trden{e} = \iota_1 \\ \sctx \vdash \trden{e_1} = \iota_1?? \\ \sctx \vdash \trden{e_2} = \iota_2} 
+{ \sctx \vdash \trden{ \rcheck{r}{e}{x.e_1}{e_2} } = \tcheck{\rx{r}}{\iota}{x.\iota_1}{\iota_2} }
+
+\end{mathpar}
+\caption{Translation from source terms ($e$) to target terms ($\iota$).
+The translation is type-directed.}
+\end{figure}
 
@@
 %\end{description}
-
-% NRF commented out section 2 in order to determine the length of section 1.
-% Replace the pattern ^% with an emptty string in the selection starting at
-% the first empty line of section 2 and ending just before \end{document}.
-
 \section{Implementation in {Atlang}}\label{atlang}
-\begin{figure}\label{fig:example}
-\begin{lstlisting}
+%\begin{figure}
+%\begin{lstlisting}
+%output of successful compilation
+%\end{lstlisting}
+%\caption{Output of successful compilation.}
+%\end{figure}
+%
+%\begin{figure}
+%\begin{lstlisting}
+%output of failed compilation
+%\end{lstlisting}
+%\caption{Output of failed compilation.}
+%\end{figure}
+%
+
+\subsection{An Example}\label{whatev}
+
+\begin{figure}[h]\begin{lstlisting}
 @fn
@@
 \vspace{-10px}
-\caption{Regular string types in atlang, a library that enables static type checking for Python. }
-\end{figure}\label{fig:impl}
+\caption{Regular string types in atlang, a library that enables static type checking for Python. }   \label{examplefig}
+
+\end{figure}
+
+Figure \ref{examplefig} demonstrates the use of string types in atlang.
+The sanitize function takes an arbitrary string and returns a string without
+double quotes or left and right brackets. In this example, we use HTML
+escape sequences.
+
+The main function receives user input and passes this input to a 
+sanitize function, which replaces all double quotes and brackets with HTML
+escape sequences.
+
+The result of applying sanitize to input is appended to two functions which
+construct a safe query and safe output.
+The arguments to the result and output construction functions constitute \emph{specifications}.
+In the case of \textsf{results_query}, this specification ensures that user input is always interpreted
+as a string literal by the SQL server.
+In the case of \textsf{results_div}, this specification ensures that user input does 
+not contain any HTML tags, which is a conservative but effective policy for preventing
+XSS attacks.
+
+Note that ${\sf input}$ does not actually meet these specifications without additional machinery.
+The type of ${\sf input}$ is quite large
+and does not actually equal the specified domains of the query or output construction
+methods. This mismatch is common -- in fact, nearly universal. Therefore, 
+our implementation includes a simple subtyping relation between regular expression
+types. 
+This subtyping relation is justified theoretically by the fact that language inclusion is
+decidable; see \cite{fulton12} for a formal definition of the subtyping relation.
+Additionally, our extension remains composable because subtyping is defined on a type-by-type basis;
+see \cite{fulton13} for a discussion of subtyping in Atlang (referred to there as Ace).
+
+\subsection{Implementation of the Regular Expression Type}
+
+\label{fig:impl}
 \begin{figure}
@@
 \end{figure}
-
-\begin{figure}
-\begin{lstlisting}
-output of successful compilation
-\end{lstlisting}
-\caption{Output of successful compilation.}
-\end{figure}
-
-\begin{figure}
-\begin{lstlisting}
-output of failed compilation
-\end{lstlisting}
-\caption{Output of failed compilation.}
-\end{figure}
-
-\subsection{An Example}
-
-Figure \ref{fig:example} demosntrates the use of string types in atlang.
-The sanitize function takes an arbitrary string and returns a string without
-double quotes or left and right brackets. In this example, we use HTML
-escape sequences.
-
-The main function receives user input and passes this input to a 
-sanitize function, which replaces all double quotes and brackets with HTML
-escape sequences.
-
-The result of applying sanitize to input is appended to two functions which
-construct a safe query and safe output.
-The arguments to the result and output construction functions constitute \emph{specifications}.
-In the case of \textsf{results_query}, this specification ensures that user input is always interpreted
-as a string literal by the SQL server.
-In the case of \textsf{results_div}, this specification ensures that user input does 
-not contain any HTML tags, which is a conservative but effective policy for preventing
-XSS attacks.
-
-Note that ${\sf input}$ does not actually meet these specifications without additional machinery.
-The type of ${\sf input}$ is quite large
-and does not actually equal the specified domains of the query or output construction
-methods. This mismatch is common -- in fact, nearly universal. Therefore, 
-our implementation includes a simple subtyping relation between regular expression
-types. 
-This subtyping relation is justified theoretically by the fact that language inclusion is
-decidable; see \cite{fulton12} for a formal definition of the subtyping relation.
-Additionally, our extension remains composable because subtyping is defined on a type-by-type basis;
-see \cite{fulton13} for a discussion of subtyping in Atlang (referred to there as Ace).
-
-\subsection{Implementation of the Regular Expression Type}
 

### 2014-08-14 sanitation-psp14/sanitation-psp14-short.tex

--- 2014-08-14 sanitation-psp14/sanitation-psp14-short.tex
@@
 
-\usepackage{todonotes}
-
+%\usepackage{todo}
+%\usepackage[subject={Todo},author={Josef}]{pdfcomment}
+%\usepackage{cooltooltips}
+\newcommand{\todo}[1]{{\color{red} #1}}
 \usepackage{placeins}
@@
 { x:\sigma \in \sctx }
-{ \sctx \vdash s:\sigma}
+{ \sctx \vdash x:\sigma}
 
@@
 {\sctx, x : \sigma_1 \vdash e : \sigma_2}
-{\sctx \vdash \lambda x:\sigma_1.e : \sigma_1 \rightarrow \sigma_2}
+{\sctx \vdash \lambda x.e : \sigma_1 \rightarrow \sigma_2}
 
@@
 
-\addtocounter{figure}{4}
 \begin{figure*}[t]\label{fig:trans}
 
-$\fbox{\inferrule{}{\trden{S} = P}}$
+$\fbox{\inferrule{}{\sctx \vdash \trden{S} = P}}$
 \begin{mathpar}
@@
 \inferrule[Tr-SafeCoerce]
-{ \ }
-{ \sctx \vdash  \trden{ \coerce{r'}{e} } = \trden{e} }
+{ \sctx \vdash \trden{e} = \iota }
+{ \sctx \vdash  \trden{ \coerce{r'}{e} } = \iota }
 
 \inferrule[Tr-Check-Ok]
-{ \sctx \vdash e : \stringin{r} \\ \lang{r} \subseteq \lang{r'} } 
-{ \sctx \vdash  \trden{ \rcheck{r'}{e}{x.e_1}{e_2} } = \trden{[e/x]e_1} }
+{ \sctx \vdash e : \stringin{r} \\ \lang{r} \subseteq \lang{r'} \\ \trden{[e/x]e_1} = \iota } 
+{ \sctx \vdash  \trden{ \rcheck{r'}{e}{x.e_1}{e_2} } = \iota }
 
 \inferrule[Tr-Check-NotOk]
-{ \sctx \vdash e : \stringin{r} \\ \lang{r} \not \subseteq \lang{r'}}
-{ \sctx \vdash  \trden{ \rcheck{r'}{e}{x.e_1}{e_2} } = \trden{\tcheck{r'}{e_2}} }
+{ \sctx \vdash e : \stringin{r} \\ \lang{r} \not \subseteq \lang{r'} \\ \trden{\tcheck{r'}{e_2}} = \iota}
+{ \sctx \vdash  \trden{ \rcheck{r'}{e}{x.e_1}{e_2} } = \iota}
 
@@
 \inferrule[P-T-Abs]
-{\tctx, x : \iota_1 \vdash e : \iota_2}
-{\tctx \vdash \lambda x:\iota_1.e : \iota_1 \rightarrow \iota_2}
+{\tctx, x : \tau_1 \vdash \iota_2 : \tau_2}
+{\tctx \vdash \lambda x.\iota_2 : \tau_1 \rightarrow \tau_2}
 
 \inferrule[P-T-App]
-{\tctx \vdash e_1 : \iota_2 \rightarrow \iota \\ \tctx \vdash e_2 : \iota_2}
-{\tctx \vdash e_1(e_2) : \iota}
+{\tctx \vdash \iota_1 : \tau_2 \rightarrow \tau \\ \tctx \vdash \iota_2 : \tau_2}
+{\tctx \vdash \iota_1(\iota_2) : \iota}
 
@@
 { e_1 \sreduces e_1' \\  e_2 \sreduces e_2' }
-{ e_1 e_2 \sreduces e_1' e_2'}
+{ e_1(e_2) \sreduces e_1'(e_2')}
 
@@
 
-\inferrule[P-E-Check-Err]
-{\iota_1 \treduces \rx{r} \\ \iota_2 \treduces \tstr{s} \\ s \not \in \lang{r}}
-{\tcheck{\iota_1}{\iota_2} \err}
 \end{mathpar}
-\caption{Big step semantics for of $\lambdap$.
-Error propagation rules are omitted.}
+\caption{Big step semantics for of $\lambdap$}.
 \label{fig:ieval}
@@
 
+@fn
 def main(db):
@@
 
-\subsection{An Example}
-
-Figure \ref{fig:atexample} demosntrates the use of string types in atlang.
+\subsection{Example Usage}
+
+Figure \ref{fig:atexample} demonstrates the use of two type constructors, \verb|fn| and \verb|string_in|, both of which we have included in \verb|atlib|, the standard library for \verb|atlang|.
+The \verb|fn| type constructor can be used to annotate functions that should be statically checked by \verb|atlang|.
+The function \verb|sanitize| on lines 3-7 specifies one argument, \verb|s|, of type \lstinline{string_in[r'.*']}.\todo{fix sizes}
+
 The sanitize function takes an arbitrary string and returns a string without

### 2014-08-18 sanitation-psp14/sanitation-psp14-short.tex

--- 2014-08-14 sanitation-psp14/sanitation-psp14-short.tex
@@
 To avoid this problem, the program must \emph{sanitize} user input.
-For example, in this case, the developer (or, more often, a framework) might define a function \verb|sanitize| that prepends double quotes (and existing backslashes)  with a backslash, which SQL treats safely.
-Guaranteeing that user input has already been sanitized before it is used to construct a command is challenging.
+For example, in this case, the developer (or, more often, a framework) might define a function \verb|sanitize| that escapes double quotes and existing backslashes   with a backslash, which SQL treats safely.
+Guaranteeing that user input has already been sanitized like this before it is used to construct a command is challenging.
 Note that this function is not idempotent, so it should only be called once.
 
-We observe that most such sanitization techniques can be understood in terms of \emph{regular languages} \cite{cinderella}.
-For example, \verb|name| should be a string in the language described by the regular expression \verb!([^"\]|(\")|(\\))*!
+We observe that many such sanitization techniques can be understood using \emph{regular languages} \cite{cinderella}.
+For example, \verb|name| must be a string in the language described by the regular expression \verb!([^"\]|(\")|(\\))*!
 -- a sequence of characters other than quotation marks and backslashes; these can only appear escaped.
-This concrete pattern syntax can be understood to desugar, in a standard way, to the syntax for regular expressions shown in Figure \ref{fig:regex}, where $r \cdot r$ is sequencing and $r + r$ is  disjunction.
+This concrete syntax for regular expression patterns can be understood to desugar, in a standard way, to the syntax for regular expressions shown in Figure \ref{fig:regex}, where $r \cdot r$ is sequencing and $r + r$ is  disjunction.
 We will work with this ``core'' for simplicity.
@@
       $\lambda x . e$ \alt
-      $e~(e)$ \alt
+      $e(e)$ \alt
       $\strin{s}$ \hfill $s \in \Sigma^{*}$ \alt
@@
 In this section, we define a minimal typed lambda calculus with regular string types called $\lambda_{RS}$.
+The syntax of $\lambda_{RS}$ is specified in Figure \ref{fig:lamrs}, its static semantics in Figure \ref{fig:etyping} and its evaluation semantics in Figure \ref{fig:eeval}.
 This will serve as the source language for our translation to a calculus with only standard strings and regular expressions, defined in Sec.
@@
 
-The syntax of $\lambda_{RS}$ is given in Figure \ref{fig:lamrs}, its static semantics are specified in Figure \ref{fig:etyping} and its evaluation semantics are specified in Figure \ref{fig:eeval}.
-
-The system has regular expression types $\stringin{r}$ whre $r$ is a regular expression. 
-Expressions of this type evaluate to string literals in the language described by $r$.
-Operations on expressions of type $\stringin{r}$ preserve this property.
+There are two type constructors in $\lambdas$, $\rightarrow$ and $\textsf{stringin}$.
+Arrow types classify functions, which are introduced via lambda abstraction, $\lambda x.e$, and can be applied, written $e(e)$, in the usual way \cite{pfpl}.
+Regular string types are of the form $\stringin{r}$, where $r$ is a regular expression.
+Values of such regular string types take the form $\strin{s}$, where $s$ is a string (i.e.
+$s \in \Sigma^\star$, defined in the usual way).
+The rule \textsc{S-T-Stringin-I} ensures that $s \in \lang{r}$.
+The remaining operations on terms of type $\stringin{r}$ preserve this property.
+%The remaining operators operate on regular strings.
+
+An explanation of significant typing rules follows:
+\begin{itemize}
+  \item \textsc{S-T-Concat}: the type of concatenated regular strings is  
+    obtained by concatenating the regular expressions.
+  \item \textsc{S-T-Replace}: \todo{fix this description}the term $\textsf{rreplace}$
+    evaluates to the result of substituting every string matching $r$ in $s_1$
+    with $s_2$. This operation corresponds to the \textsf{str_replace} function of PHP.
+    The type of these expressions is defined in terms of an extra-linguistic \textsf{lreplace}
+    function, which is defined later in this section.
+    The expression $\lreplace{r}{r_1}{r_2}$ is obtained by
+    starting with $r_1$, and replacing any subexpression matching $r$ with $r_2$.
+  \item \textsc{S-T-SafeCoerce}: \todo{fix this description} allows only safe coercions between string types by
+    expoiting the decidability of language inclusion.
+  \item \textsc{S-T-Check}: \todo{fix this description}The check rules allow casts between strings that cannot be guaranteed at compile time, but might pass a runtime check.
+  \item \todo{strcase}
+\end{itemize}
 
@@
 \end{itemize}
+
+\begin{figure}\label{fig:lambdas}
+\small
+$\fbox{\inferrule{}{\sctx \vdash e : \sigma}}$
+~~~~$\sctx ::= \emptyset \pipe \sctx, x : \sigma$
+\begin{mathpar}
+\inferrule[S-T-Var]
+{ x:\sigma \in \sctx }
+{ \sctx \vdash x:\sigma}
+
+\inferrule[S-T-Abs]
+{\sctx, x : \sigma_1 \vdash e : \sigma_2}
+{\sctx \vdash \lambda x.e : \sigma_1 \rightarrow \sigma_2}
+
+\inferrule[S-T-App]
+{\sctx \vdash e_1 : \sigma_2 \rightarrow \sigma \\ \sctx \vdash e_2 : \sigma_2}
+{\sctx \vdash e_1(e_2) : \sigma}
+
+\inferrule[S-T-Stringin-I]
+{s \in \lang{r}}
+{\sctx \vdash \strin{s} : \stringin{r}}
+
+\inferrule[S-T-Concat]
+{\sctx \vdash e_1 : \stringin{r_1} \\ \sctx \vdash e_2 : \stringin{r_2}}
+{\sctx \vdash \rsconcat{e_1}{e_2} : \stringin{r_1 \cdot r_2}}
+
+\inferrule[S-T-Replace]
+{\sctx \vdash e_1 : \stringin{r_1} \\ \sctx \vdash e_2 : \stringin{r_2} \\\\ \lreplace{r}{r_1}{r_2} = r'}
+{\sctx \vdash \sisubst{r}{e_1}{e_2} : \stringin{r'}}
+
+\inferrule[S-T-SafeCoerce]
+{\sctx \vdash e : \stringin{r'} \\ \lang{r'} \subseteq \lang{r}}
+{\sctx \vdash \coerce{r}{e} : \stringin{r}}
+
+\inferrule[S-T-Check]
+{\sctx \vdash e_0 : \stringin{r_0} \\ \sctx, x:\stringin{r} \vdash e_1 : \sigma \\ \sctx \vdash e_2 : \sigma}
+{\sctx \vdash \rcheck{r}{e_0}{x.e_1}{e_2} : \sigma}
+
+\end{mathpar}
+
+\caption{Typing rules for $\lambdas$. 
+The typing context $\sctx$ is standard.}
+\label{fig:etyping}
+\end{figure}
+
+\begin{figure}
+\small
+$\fbox{\inferrule{}{e \sreduces e}}$
+\begin{mathpar}
+\inferrule[S-E-Abs]                                                             
+{ \ }                                                                           
+{\lambda x.e \sreduces \lambda x.e}
+
+\inferrule[S-E-App]
+{ e_1 \sreduces e_1' \\ e_2 \sreduces e_2' }
+{ e_1(e_2) \sreduces e_1'(e_2')}
+
+\inferrule[S-E-RStr]
+{ \ }
+{\strin{s} \sreduces \strin{s}}
+
+\inferrule[S-E-Concat]
+{e_1 \sreduces \strin{s_1} \\ e_2 \sreduces \strin{s_2}} 
+{\rsconcat{e_1}{e_2} \sreduces \strin{s_1 s_2}} % ???
+
+\inferrule[S-E-Replace]
+{e_1 \sreduces \strin{e_1} \\ e_2 \sreduces \strin{s_2} \\ \lsubst{r}{s_1}{s_2} = s} 
+{\sisubst{r}{e_1}{e_2} \sreduces \sistr{s}}
+
+\inferrule[S-E-SafeCoerce]
+{e \sreduces \strin{s}}
+{\coerce{r}{e} \sreduces \strin{s}}
+
+\inferrule[S-E-Check-Ok]
+{e \sreduces \strin{s} \\ s \in \lang{r}}
+{
+  \rcheck{r}{e}{x.e_1}{e_2} \sreduces [\strin{s} / x]e_1
+}
+
+\inferrule[S-E-Check-NotOk]
+{
+  e \sreduces \strin{s} \\ s \not \in \lang{r}
+}
+{
+  \rcheck{r}{e}{x.e_1}{e_2} \sreduces e_2
+}
+
+\end{mathpar}
+\caption{Big step semantics for $\lambdas$}.
+\label{fig:eeval}
+\end{figure}
 
@@
 The $\lambdas$ system extends the simply-typed lambda calculus with regular
-expression types; an explanation of significant typing rule follows:
-\begin{itemize}
-  \item Rule S-T-Concat states that the type of concatenated strings is 
-    obtained by concatenating the regular expressions for each string.
-  \item Rule S-T-Replace defines the type of $\textsf{rreplace}$. The expression $\textsf{rreplace}$
-    evaluates to the result of substituting every string matching $r$ in $s_1$
-    with $s_2$. This operation corresponds to the \textsf{str_replace} function of PHP.
-    The type of these expressions is defined in terms of an extra-linguistic \textsf{lreplace}
-    function, which is defined later in this section.
-    The expression $\lreplace{r}{r_1}{r_2}$ is obtained by
-    starting with $r_1$, and replacing any subexpression matching $r$ with $r_2$.
-  \item Rule S-T-SafeCoerce allows only safe coercions between string types by
-    expoiting the decidability of language inclusion.
-  \item The check rules allow casts between strings that cannot be guaranteed at compile time, but might pass a runtime check.
-\end{itemize}
-
-\newpage
+expression types
+
+%\newpage
 
@@
 \label{fig:ieval}
-\end{figure}
-
-\begin{figure}\label{fig:lambdas}
-\small
-$\fbox{\inferrule{}{\sctx \vdash e : \sigma}}$
-~~~~$\sctx ::= \emptyset \pipe \sctx, x : \sigma$
-\begin{mathpar}
-\inferrule[S-T-Var]
-{ x:\sigma \in \sctx }
-{ \sctx \vdash x:\sigma}
-
-\inferrule[S-T-Abs]
-{\sctx, x : \sigma_1 \vdash e : \sigma_2}
-{\sctx \vdash \lambda x.e : \sigma_1 \rightarrow \sigma_2}
-
-\inferrule[S-T-App]
-{\sctx \vdash e_1 : \sigma_2 \rightarrow \sigma \\ \sctx \vdash e_2 : \sigma_2}
-{\sctx \vdash e_1(e_2) : \sigma}
-
-\inferrule[S-T-Stringin-I]
-{s \in \lang{r}}
-{\sctx \vdash \strin{s} : \stringin{r}}
-
-\inferrule[S-T-Concat]
-{\sctx \vdash e_1 : \stringin{r_1} \\ \sctx \vdash e_2 : \stringin{r_2}}
-{\sctx \vdash \rsconcat{e_1}{e_2} : \stringin{r_1 \cdot r_2}}
-
-\inferrule[S-T-Replace]
-{\sctx \vdash e_1 : \stringin{r_1} \\ \sctx \vdash e_2 : \stringin{r_2} \\\\ \lreplace{r}{r_1}{r_2} = r'}
-{\sctx \vdash \sisubst{r}{e_1}{e_2} : \stringin{r'}}
-
-\inferrule[S-T-SafeCoerce]
-{\sctx \vdash e : \stringin{r'} \\ \lang{r'} \subseteq \lang{r}}
-{\sctx \vdash \coerce{r}{e} : \stringin{r}}
-
-\inferrule[S-T-Check]
-{\sctx \vdash e_0 : \stringin{r_0} \\ \sctx, x:\stringin{r} \vdash e_1 : \sigma \\ \sctx \vdash e_2 : \sigma}
-{\sctx \vdash \rcheck{r}{e_0}{x.e_1}{e_2} : \sigma}
-
-\end{mathpar}
-
-\caption{Typing rules for $\lambdas$. 
-The typing context $\sctx$ is standard.}
-\label{fig:etyping}
-\end{figure}
-
-\begin{figure}
-\small
-$\fbox{\inferrule{}{e \sreduces e}}$
-\begin{mathpar}
-\inferrule[S-E-Abs]                                                             
-{ \ }                                                                           
-{\lambda x.e \sreduces \lambda x.e}
-
-\inferrule[S-E-App]
-{ e_1 \sreduces e_1' \\ e_2 \sreduces e_2' }
-{ e_1(e_2) \sreduces e_1'(e_2')}
-
-\inferrule[S-E-RStr]
-{ \ }
-{\strin{s} \sreduces \strin{s}}
-
-\inferrule[S-E-Concat]
-{e_1 \sreduces \strin{s_1} \\ e_2 \sreduces \strin{s_2}} 
-{\rsconcat{e_1}{e_2} \sreduces \strin{s_1 s_2}} % ???
-
-\inferrule[S-E-Replace]
-{e_1 \sreduces \strin{e_1} \\ e_2 \sreduces \strin{s_2} \\ \lsubst{r}{s_1}{s_2} = s} 
-{\sisubst{r}{e_1}{e_2} \sreduces \sistr{s}}
-
-\inferrule[S-E-SafeCoerce]
-{e \sreduces \strin{s}}
-{\coerce{r}{e} \sreduces \strin{s}}
-
-\inferrule[S-E-Check-Ok]
-{e \sreduces \strin{s} \\ s \in \lang{r}}
-{
-  \rcheck{r}{e}{x.e_1}{e_2} \sreduces [\strin{s} / x]e_1
-}
-
-\inferrule[S-E-Check-NotOk]
-{
-  e \sreduces \strin{s} \\ s \not \in \lang{r}
-}
-{
-  \rcheck{r}{e}{x.e_1}{e_2} \sreduces e_2
-}
-
-\end{mathpar}
-\caption{Big step semantics for $\lambdas$}.
-\label{fig:eeval}
 \end{figure}
@@
 \begin{mathpar}
+\inferrule[Tr-Var]
+{ }
+{\sctx \vdash \trden{x} = x}
 
@@
 
+\section{Implementation in {Atlang}}\label{atlang}
+In the previous section, we specified a type system and a translation semantics to a language containing only strings and regular expressions.
+In this section, we take Python to be such a target language.
+Python does not have a static type system, however, so to implement these semantics, we will leverage \verb|atlang|, an extensible type system for Python (being developed by the authors).
+By using \verb|atlang|, which builds on Python's quotations and reflection facilities, we can implement these semantics as a library, rather than as a new dialect of Python.
+
 \begin{figure}[h]\begin{lstlisting}
+from atlib import fn, string_in
+
 @fn
@@
 \vspace{-10px}
-\caption{Regular string types in atlang, a library that enables static type checking for Python. }   \label{examplefig}
-
+\caption{Regular string types in atlang, a library that enables static type checking for Python. }\label{fig:atexample}
 \end{figure}
@@
 \caption{Implementation of the \texttt{string\_in} type constructor in atlang.}
-\end{figure}
+\label{fig:atimpl}\end{figure}
+
+\subsection{Example Usage}
+
+Figure \ref{fig:atexample} demonstrates the use of two type constructors, \verb|fn| and \verb|string_in|, both of which we have included in \verb|atlib|, the standard library for \verb|atlang|.
+The \verb|fn| type constructor can be used to annotate functions that should be statically checked by \verb|atlang|.
+The function \verb|sanitize| on lines 3-7 specifies one argument, \verb|s|, of type \lstinline{string_in[r'.*']}.\todo{fix sizes}
+
+The sanitize function takes an arbitrary string and returns a string without
+double quotes or left and right brackets. In this example, we use HTML
+escape sequences.
+
+The main function receives user input and passes this input to a 
+sanitize function, which replaces all double quotes and brackets with HTML
+escape sequences.
+
+The result of applying sanitize to input is appended to two functions which
+construct a safe query and safe output.
+The arguments to the result and output construction functions constitute \emph{specifications}.
+In the case of \textsf{results_query}, this specification ensures that user input is always interpreted
+as a string literal by the SQL server.
+In the case of \textsf{results_div}, this specification ensures that user input does 
+not contain any HTML tags, which is a conservative but effective policy for preventing
+XSS attacks.
+
+Note that ${\sf input}$ does not actually meet these specifications without additional machinery.
+The type of ${\sf input}$ is quite large
+and does not actually equal the specified domains of the query or output construction
+methods. This mismatch is common -- in fact, nearly universal. Therefore, 
+our implementation includes a simple subtyping relation between regular expression
+types. 
+This subtyping relation is justified theoretically by the fact that language inclusion is
+decidable; see \cite{fulton12} for a formal definition of the subtyping relation.
+Additionally, our extension remains composable because subtyping is defined on a type-by-type basis;
+see \cite{fulton13} for a discussion of subtyping in Atlang (referred to there as Ace).
+
+\subsection{Implementation of the Regular Expression Type}
 
@@
 
-\subsection{Related Work and Alternative Approaches}
+\section{Related Work and Alternative Approaches}\label{related}
 

### 2014-08-18 sanitation-psp14/sanitation-psp14-short.tex
Monday meeting changes #1
 * Fixed semantics (audit the check rules)
 * Fixed names of thms/lems/defs and changed lems to props in sec 2

@@
 \inferrule[S-E-App]
-{ e_1 \sreduces e_1' \\ e_2 \sreduces e_2' }
-{ e_1(e_2) \sreduces e_1'(e_2')}
+{ e_1 \sreduces \lambda x . e_3 \\ e_2 \sreduces e_2' \\ [ e_2' / x ] e_3 \sreduces v }
+{ e_1(e_2) \sreduces v }
 
@@
 \inferrule[S-E-Check-Ok]
-{e \sreduces \strin{s} \\ s \in \lang{r}}
+{e \sreduces \strin{s} \\ s \in \lang{r} \\ [\strin{s} / x]e_1 \sreduces v}
 {
-  \rcheck{r}{e}{x.e_1}{e_2} \sreduces [\strin{s} / x]e_1
+  \rcheck{r}{e}{x.e_1}{e_2} \sreduces v
 }
@@
 \inferrule[P-E-App]
-{ \iota_1 \sreduces \iota_1' \\  \iota_2 \sreduces \iota_2' }
-{ \iota_1(\iota_2) \sreduces \iota_1'(\iota_2')}
+{ \iota_1 \sreduces \lambda x . \iota_3 \\  \iota_2 \sreduces \iota_2' \\ [\iota_3 / x] \iota_2' \sreduces v}
+{ \iota_1(\iota_2) \sreduces v}
 
@@
 \inferrule[P-E-Check-OK]
-{\iota \treduces \tstr{s} \\ s \in \lang{r}}
-{\tcheck{\rx{r}}{\iota}{x.\iota_1}{\iota_2} \treduces [\tstr{s}/x]\iota_1}
+{\iota \treduces \tstr{s} \\ s \in \lang{r} \\ [\tstr{s} / x] \iota_1 \sreduces \iota_3}
+{\tcheck{\rx{r}}{\iota}{x.\iota_1}{\iota_2} \treduces \iota_3}
 
@@
 
-\begin{lem}{Properties of Regular Languages and Expressions.} \label{thm:regexprops}
+\begin{lem}[Properties of Regular Languages and Expressions.] \label{thm:regexprops}
 The following are properties of regular expressions which are necessary for our proofs:
@@
 
-\begin{lem}{Closure.} \label{thm:total}
+\begin{prop}[Closure.] \label{thm:total}
   If $\lang{r}, \lang{r_1}$ and $\lang{r_2}$ are regular expressions, then $\lang{\lreplace{r}{r_1}{r_2}}$ is also a regular language.
-\end{lem}
+\end{prop}
 \begin{proof}
@@
 
-\begin{lem}{Substitution Correspondence.} \label{thm:substcorrespondence}
+\begin{prop}[Substitution Correspondence.] \label{thm:substcorrespondence}
   If $s_1 \in \lang{r_1}$ and $s_2 \in \lang{r_2}$ then $\lsubst{r}{s_1}{s_2} \in \lang{\lreplace{r}{s_1}{s_2}}$.
-\end{lem}
+\end{prop}
 \begin{proof}
@@
 
-\begin{thm}{Type Safety for the String Fragment of $\lambdas$.} \label{thm:scorrect}
-  Let $S$ be a term in the source language. If $\sctx \vdash e : \stringin{r}$ 
+\begin{thm}[Type Safety for the String Fragment of $\lambdas$.] \label{thm:scorrect}
+  If $\sctx \vdash e : \stringin{r}$ 
   then $e \sreduces \rstr{s}$ and $\sctx \vdash \rstr{s} : \stringin{r}$.
@@
 
-\begin{thm}{Correctness of Input Sanitation for $\lambdas$.}\label{thm:scorrect}
+\begin{thm}[Correctness of Input Sanitation for $\lambdas$.]\label{thm:scorrect}
   If  $\sctx \vdash e : \stringin{r}$ and $e \sreduces \rstr{s}$ then $s \in \lang{r}$.
@@
 
-\begin{thm}{Translation Correctness.} \label{thm:tcorrect}
+\begin{thm}[Translation Correctness.] \label{thm:tcorrect}
 If $\sctx \vdash e : \stringin{r}$ then there exists a $\iota$ such that $\trden{e} = \iota$, 
@@
 
-\begin{thm}{Correctness of Input Sanitation for Translated Terms.}\label{thm:main}
+\begin{thm}[Correctness of Input Sanitation for Translated Terms.]\label{thm:main}
   If $\trden{e} = \iota$ and $\sctx \vdash e : \stringin{r}$ then $\iota \sreduces \tstr{s}$

### 2014-08-18 sanitation-psp14/sanitation-psp14-short.tex
Finished trivial edits.

@@
 \begin{thm}[Type Safety for the String Fragment of $\lambdas$.] \label{thm:scorrect}
-  If $\sctx \vdash e : \stringin{r}$ 
-  then $e \sreduces \rstr{s}$ and $\sctx \vdash \rstr{s} : \stringin{r}$.
+  If $\sctx \vdash e : \sigma$ 
+  then $e \sreduces e'$ and $\sctx \vdash e' : \stringin{r}$.
 \end{thm}
@@
 
+\begin{thm}[Canonical Forms for String Fragment of $\lambdas$.]\label{thm:cfs}
+  If $\sctx \vdash e : \stringin{r}$ 
+  then $e \sreduces \rstr{s}$ 
+\end{thm}
+
 In addition to safety, $\lambdas$ requires a correctness result ensuring that well-typed terms of regular string type are in the language associated with their type.
@@
 \begin{proof}
-  Follows directly from type safety, canonical forms for $\lambdas$.
+  Follows directly from type safety, canonical forms for $\lambdas$, and inversion of the typing relation for $\lambdas$.
 \end{proof}
@@
 \begin{thm}[Translation Correctness.] \label{thm:tcorrect}
-If $\sctx \vdash e : \stringin{r}$ then there exists a $\iota$ such that $\trden{e} = \iota$, 
+If $\sctx \vdash e : \stringin{r}$ then there exists an $\iota$ such that $\trden{e} = \iota$, 
 $\iota \treduces \tstr{s}$, and $e \sreduces \strin{s}$.

### 2014-08-18 sanitation-psp14/sanitation-psp14-short.tex
strcase

@@
 % Meta-theoretic functions
-\newcommand{\lsubst}[3]{{\tt subst}(#1;#2;#3)} % This used to renderlreplace(...) so there're probably mistkes wherever this command was used now.
-\newcommand{\lreplace}[3]{{\tt lreplace}(#1; #2; #3)}
+\newcommand{\lsubst}[3]{{\sf subst}(#1;#2;#3)} % This used to renderlreplace(...) so there're probably mistkes wherever this command was used now.
+\newcommand{\lreplace}[3]{{\sf lreplace}(#1; #2; #3)}
+
+\newcommand{\strcase}[3]{ {\sf strcase}(#1; #2; #3)}
+\newcommand{\pstrcase}[3]{ {\sf pstrcase}(#1; #2; #3)}
+
+\newcommand{\lhead}[1]{ {\sf lhead}(#1) }
+\newcommand{\ltail}[1]{ {\sf ltail}(#1) }
 
@@
 
+\begin{defn}[Definition of \lhead{r}]
+The function $\lhead{r}$ is defined in terms of the structure of $r$:
+\begin{itemize}
+\item $\lhead{a r'} = a$ where $a \in \Sigma$
+\item $\lhead{r_1 \cup  r_2}  = \lhead{r_1} \cup \lhead{r_2}$
+\item $\lhead{r*} = \lhead{r}$
+\item $\lhead{.} = a_1 \cup a_2 \cup ... \cup a_n$ for all $a_i \in \Sigma$ where $|\Sigma| = n$
+\item $\lhead{\epsilon} = \epsilon$.
+\end{itemize}
+\end{defn}
+
+\begin{defn}[Definition of \ltail{r}]
+The function \ltail{r} is defined in terms of $\lhead{r}$.
+Note that $\lhead{r} = a_1 \cup a_2 \cup ... \cup a_i$.
+We define $\ltail{r} = \delta_{a_1}(r) \cup \delta_{a_2}(r) \cup ... \cup \delta_{a_i}(r)$
+where $\delta_s(r)$ is the Brzozowski derivative of $r$ with respect to $s$ \cite{bowzer}.
+\end{defn}
+
 The premier operation for manipulating strings in $\lambdas$ is string substitution, 
@@
 
+\inferrule[S-T-Case]
+{ \sctx \vdash e_1 : \stringin{r} \\
+  \sctx \vdash e_2 : \sigma \\
+  \sctx, x : \stringin{\lhead{r}}, y : \stringin{\ltail{r}} \vdash e_3 : \sigma
+}
+{
+  \sctx \vdash \strcase{e_1}{e_2}{x.y.e_3} : \sigma
+}
+
 \inferrule[S-T-Replace]
@@
 {\rsconcat{e_1}{e_2} \sreduces \strin{s_1 s_2}} % ???
+
+\inferrule[S-E-Case-$\epsilon$]
+{
+  e_1 \sreduces \strin{\epsilon} \\
+  e_2 \sreduces v_2
+}
+{
+  \strcase{e_1}{e_2}{e_3} \sreduces v_2
+}
+
+\inferrule[S-E-Case-Concat]
+{
+  e_1 \sreduces \strin{p s} \\
+  e_3 \sreduces x.y.e_4 \\
+  [p/x][s/y]e_4 \sreduces v
+}
+{
+  \strcase{e_1}{e_2}{e_3} \sreduces v
+}
 
@@
 
+\inferrule[P-T-Case]
+{\tctx \vdash \iota_1 : \str \\ \tctx \vdash \iota_2 : \tau \\ \tctx, x:\str, y:\str \vdash \iota_3:\tau}
+{\tctx \vdash \pstrcase{\iota_1}{\iota_2}{\iota_3} : \tau}
+
 \inferrule[P-T-Replace]
@@
 {\tconcat{\iota_1}{\iota_2} \treduces \tstr{s_1 s_2}} % ???
+
+\inferrule[P-E-Case-$\epsilon$]
+{
+  \iota_1 \treduces \tstr{\ } \\
+  \iota2 \treduces \iota_2
+}
+{
+  \pstrcase{\iota_1}{\iota_2}{\iota_3} \treduces v_2
+}
+
+\inferrule[P-E-Case-Concat]
+{
+  \iota_1 \treduces \tstr{ps} \\
+  x.y.\iota_3 \treduces \iota_4 \\
+  [p/x][s/y]\iota_4 \treduces v
+}
+{
+  \pstrcase{\iota_1}{\iota_2}{\iota_3} \treduces v
+}
 
@@
 {\sctx \vdash \trden{e_1(e_2)} = \iota_1(\iota_2)}
+
+\inferrule[Tr-Case]
+{\sctx \vdash \trden{e_1} = \iota_1 \\
+  \sctx \vdash \trden{e_2} = \iota_2 \\
+  \sctx \vdash \trden{e_3} = \iota_3
+}
+{\sctx \vdash \trden{ \strcase{e_1}{e_2}{e_3} } = \pstrcase{\iota_1}{\iota_2}{\iota_3} }
 

### 2014-08-19 sanitation-psp14/sanitation-psp14-short.tex
Added new structure. Beginning copy-paste session.
Finished reorganization, but the diagrams are a mess.

@@
 
-An explanation of significant typing rules follows:
+An outline for this section follows:
 \begin{itemize}
-  \item \textsc{S-T-Concat}: the type of concatenated regular strings is  
-    obtained by concatenating the regular expressions.
-  \item \textsc{S-T-Replace}: \todo{fix this description}the term $\textsf{rreplace}$
-    evaluates to the result of substituting every string matching $r$ in $s_1$
-    with $s_2$. This operation corresponds to the \textsf{str_replace} function of PHP.
-    The type of these expressions is defined in terms of an extra-linguistic \textsf{lreplace}
-    function, which is defined later in this section.
-    The expression $\lreplace{r}{r_1}{r_2}$ is obtained by
-    starting with $r_1$, and replacing any subexpression matching $r$ with $r_2$.
-  \item \textsc{S-T-SafeCoerce}: \todo{fix this description} allows only safe coercions between string types by
-    expoiting the decidability of language inclusion.
-  \item \textsc{S-T-Check}: \todo{fix this description}The check rules allow casts between strings that cannot be guaranteed at compile time, but might pass a runtime check.
-  \item \todo{strcase}
+\item In section \ref{sec:rs}, we describe $\lambdas$. We also prove provide proof outlines for type soundness and a correctness theorem.
+\item In section \ref{sec:p}, we present a simple target language with a minimal regular expression library.
+\item Section \ref{sec:tr} describes the translation from $\lambdas$ to $\lambdap$ and ensures the correcness result for \ref{sec:rs} is preserved under translation. 
 \end{itemize}
+
+\subsection{The Language of Constrained Strings}\label{sec:rs}
+
+The system $\lambdas$ provides many familiar operations on strings.
+However, unlike most languages, the type system of $\lambdas$ tracks the effect of these operations by relating computations on strings to corresponding computations on regular languages.
+This section describes our approach, culminating in a Security Theorem.
+The main technical result of this paper will show that this Security Theorem is preserved when $\lambdas$ is translated to a more traditional language.
+Throughout this section, we refer to \textsc{Inference Rules} defined in the figures contained in this section.
+\todo{remove tranditional, most and improve description}
+
+\subsubsection{Concatenation and String Decomposition}
+
+The \textsc{S-T-Concat} rule is the simplest example of our approach.
+The rule is sound because the result of concatenating two strings will always be in the sequential composition two regular expressions matching the respective strings.
+The rule therefore relates string concatenation to sequential composition of regular expressions.
+
+Whereas concatenation allows the construction of large strings from smaller strings, \textsc{S-T-Case} allows the decomposition, or elimination, of large strings from smaller strings.
+Intuitively, this rule ``peels off" the first character, and then performs a specified operation on the first character and the remaining string.
+In terms of strings, this is essentially the ubiquitious substring operation.
+
+The definition of \textsc{S-T-Case} is subtle because even simple string operations sometimes require complicated operations on regular expressions.
+Peeling off the first character of a string is deterministic -- the first character is always one, concrete value.
+However, a regular expression describing the first character of any matching string might recognize multiple concrete values.
+For instance, the first character of ``CMU" is always ``C"; however, if the expression associated with this string is CMU$\cup$KIT, then the expression describing the first character is C$\cup$K.
+
+Fortunately, regular expression derivatives \cite{bowzer} convienantly capture this intuition.
+The regular expression recognizing any one-character prefix of the strings in $\lang{r}$ is easily defined.
+The one-character prefix definition combined with derivatives gives a concise definition of the
+remaining string.
 
@@
 
+Given this definition of $\lhead{r}$, regular expression derivatives provide a useful tool for defining $\ltail{r}$.
+
 \begin{defn}[Definition of \ltail{r}]
@@
 
-The premier operation for manipulating strings in $\lambdas$ is string substitution, 
-which is a familiar operation to any programmer who has used regular expressions.
-The replacement operation replaces all instances of a pattern in one string with
-another string; for instance, $\lsubst{a|b}{a}{c} = c$.
-In order to compute the type resulting from substitution, we also need to compute
-the result of replacing one language with another inside a given language.
-Finally, just for convenience, we provide a coerce operation. The introduction of
-coercion requires handling of runtime errors.
-
-The underlying language $\lambdap$ has only one type for strings. We prove that whenever
-a term is translated from $\lambdas$ to $\lambdap$, correctness is preserved.
-The only exception is in the case of unsafe casts in $\lambdas$, which are unnecessary but
-are included to demonstrate that the regex library of $\lambdap$ may be used to insert
-dynamic checks whenever even when developers are not careful about using statically checked
-operations.
-
-A brief outline of this section follows:
-\begin{itemize}
-  \item In \S 2.1-2.3, we define $\lambdas$, $\lambdap$ and the translation from one to the other.
-  \item In \S 2.4 we state some properties about regular expressions which are needed in our correctness proofs.
-   \item In \S 2.5 we prove type safety for $\lambdap$ as well as both type safety and correctness for $\lambdas$.
-  \item In \S 2.6 we prove that translation preserves the correctness result about $\lambdas$.
-\end{itemize}
+The \textsc{S-T-Concat} rule, which is defined in terms of these operations, relates the result of ``peeling off" the first character of a string to regular expression derivatives.
+
+\subsubsection{Coercion}
+
+ \textsc{S-T-SafeCoerce}: \todo{fix this description} allows only safe coercions between string types by
+    expoiting the decidability of language inclusion.
+
+\textsc{S-T-Check}: \todo{fix this description}The check rules allow casts between strings that cannot be guaranteed at compile time, but might pass a runtime check.
+
+\subsubsection{Replacement}
+
+The premier operation for manipulating strings in $\lambdas$ is string substitution.
+On strings, the string substitution operator is familiar, and roughly analogous to \lstinline{str_replace} in PHP and \lstinline{String.replace} in Java.
+The definition of \textsc{S-T-Case} requires a definition of replacement for regular expressions (or languages) corresponding to string replacement.
+In this section, we define this operation.
+
+Both string and language replacement are defined extra-linguistically.
+The system $\lambdas$ is defined in terms of these functions.
+The function $\lsubst{r}{s_1}{s_2}$ replaces every substring of $s_1$ matching $r$ with $s_2$.
+In an analogous manner, the function $\lreplace{r}{r_1}{r_2}$ replaces every sublanguages of $r_1$ matching $r$ with $r_2$.
+If an intuiition for this operation is not clear, it may be helpful to think in terms replacing sub-automata.
+
+Throughout this section, we fix an alphabet $\Sigma$ over which strings $s$ and
+regular expressions $r$ are defined. We use $\lang{r}$ to refer to the
+language recognized by the regular expression $r$.
+
+\begin{lem}[Properties of Regular Languages and Expressions.] \label{thm:regexprops}
+The following are properties of regular expressions which are necessary for our proofs:
+If $s_1 \in \lang{r_1}$ and $s_2 \in \lang{r_2}$ then $s_1s_2 \in \lang{r_1r_2}$.
+For all strings $s$ and regular expressions $r$, either $s \in \lang{r}$ or $s \not \in \lang{r}$.
+Regular languages are closed under difference, right quotient, reversal, and string homomorphism.
+\end{lem}
+
+If any of these properties are unfamiliar, the reader may refer to a standard text on the subject \cite{cinderella}.
+
+\begin{defn}[$\tt{subst}$]
+The relation $\lsubst{r}{s_1}{s_2} = s$ produces a string $s$ in which all substrings of $s_1$ matching $r$ are replaced with $s_2$.
+\end{defn}
+
+\begin{defn}[$\tt{lreplace}$]
+  The relation $\lreplace{r}{r_1}{r_2} = r'$ relates $r, r_1,$ and $r_2$ to
+  a language $r'$ containing all strings of $r_1$ except that any substring $s_{pre} s s_{post} \in \lang{r_1}$ where $s \in \lang{r}$
+  is replaced by the set of strings $s_{pre} s_2 s_{post}$ for all $s_2 \in \lang{r_2}$ (the prefix and postfix positions may be empty).
+\end{defn}
+
+\begin{prop}[Closure.] \label{thm:total}
+  If $\lang{r}, \lang{r_1}$ and $\lang{r_2}$ are regular expressions, then $\lang{\lreplace{r}{r_1}{r_2}}$ is also a regular language.
+\end{prop}
+\begin{proof}
+The theorem follows from closure under difference, right quotient, reversal and string homomorphism.
+\end{proof}
+
+\begin{prop}[Substitution Correspondence.] \label{thm:substcorrespondence}
+  If $s_1 \in \lang{r_1}$ and $s_2 \in \lang{r_2}$ then $\lsubst{r}{s_1}{s_2} \in \lang{\lreplace{r}{s_1}{s_2}}$.
+\end{prop}
+\begin{proof}
+The theorem follows from the definitions of subst and lreplace; note that language substitutions over-approximate string substitutions.
+\end{proof}
+
+\subsubsection{Safety}
+
+Type soundness for $\lambdas$ follows from the lemmas and propositions established in the previous subsection.
+
+In this section, we establish type soundness for $\lambdas$.
+The theorem relies upon the lemmas and propositions established above.
+We also rely on additional lemmas which establish the preservation of well-formedness
+of regular epxressions.
+
+\begin{lem}
+  If $\sctx \vdash e : \stringin{r}$ then $r$ is a well-formed regular expression.
+\end{lem}
+\begin{proof}
+  The only non-trivial case is S-T-Replace, which follows from lemma \ref{thm:total}. 
+\end{proof}
+\begin{lem}
+If $\tctx \vdash \iota : \regex$ then $\iota \treduces \rx{r}$ such that $r$ is a well-formed regular expression. 
+\end{lem}
+
+Safety for the string fragment of $\lambdas$ requires validating that the type system's definition is justified
+by our theorems about regular languages.
+
+\begin{thm}[Canonical Forms for String Fragment of $\lambdas$.]\label{thm:cfs}
+  If $\sctx \vdash e : \stringin{r}$ 
+  then $e \sreduces \rstr{s}$ 
+\end{thm}
+
+\begin{thm}[Type Safety.] \label{thm:scorrect}
+  If $\sctx \vdash e : \sigma$ 
+  then $e \sreduces e'$ and $\sctx \vdash e' : \stringin{r}$.
+\end{thm}
+\begin{proof}
+By induction on the typing relation.
+The S-T-Concat case requires Lemma \ref{thm:regexprops} 
+and the S-T-Replace case appeals to Lemma \ref{thm:substcorrespondence}.
+\end{proof}
+
+\subsubsection{The Security Theorem}\label{sec:securitythm}
+
+The chief benefit of $\lambdas$ is its safety theorem, which states that any value of
+a regular expression type is recognized by the regular expression corresponding to its type.
+Our main technical result, stated later in this section, establishes that this property
+is preserved under translation into $\lambdap$.
+
+\begin{thm}[Correctness of Input Sanitation for $\lambdas$.]\label{thm:scorrect}
+  If  $\sctx \vdash e : \stringin{r}$ and $e \sreduces \rstr{s}$ then $s \in \lang{r}$.
+\end{thm}
+\begin{proof}
+  Follows directly from type safety, canonical forms for $\lambdas$, and inversion of the typing relation for $\lambdas$.
+\end{proof}
+
+\subsection{A Target Language with a Regular Expression Library}\label{sec:p}
+
+The system $\lambdap$ is a straight-forward extension of a simply typed lambda calculus
+with a $\str$ type and a regular expression type, as well as 
+some operations found in any standard library such as string concatenation and replacement.
+The operations of $\lambdap $ correspond precisely to the operations of $\lambdas$ in a way made precise by the translation rules described in the next section.
+Unlike $\lambdas$, $\lambdap$ does not statically track the effects of string operations.
+The language $\lambdap$ is so-called because it is reminscent of popular web programming languages,
+such as \textbf{P}ython or \textbf{P}HP.
+
+Type safety for $\lambdap$ is straight-forward, but is necessary in order to establish the 
+correctness of our translation.
+
+\begin{thm} 
+  Let $\iota$ be a term in the target language. If $\tctx \vdash \iota : \tau$ 
+  then $\iota \sreduces \iota'$ and $\tctx \vdash \iota' : \tau$.
+\end{thm}
+
+\subsection{Translation from $\lambdas$ to $\lambdap$}\label{sec:tr}
+
+The translation from $\lambdas$ to $\lambdap$ is defined in figure 5.
+The coercion cases are most interesting. If the safety of coercion in manifest in the
+types of the expressions, then no runtime check is inserted.
+If the safety of coercion is not manifest in the types, then a check is inserted.
+
+In practice, the type of a replacement rarely matches a specification.
+Therefore, it is convienant in an implementation to always insert the appropriate
+safe coercion. This way, violations of the sanitation protocol manifest as type errors,
+but the definition of an adequate replacement is not arduous.
+Alternatively, this policy may 
+be codified in the type system itself using subtyping \cite{fulton12}.
+
+This section ultimately establishes that the security theorem for $\lambdas$ is preserved under compilation.
+
+\begin{thm}[Translation Correctness]\label{thm:trcorrect}
+  If $\tctx \vdash e : \stringin{r}$ then there exists an $\iota$ such that $\trden{e} = \iota$, $\iota \sreduces \tstr{s}$,
+  and $e \treduces \rstr{s}$.
+\end{thm}
+\begin{proof}
+The proof proceeds by induction on the typing relation for $e$ and an appropraite choich of $\iota$; in each case,
+the choice is the syntactic form in $\lambdap$ corresponding to the form under consideration (e.g. choose {\sf preplace} when considering {\sf sreplace}).
+After the correct choice, the proof proceeds by our type safety theorems and an appeal to the induction hypothesis.
+\end{proof}
+
+Finally, our main result establishes that correctness of $\lambdas$ is preserved under the translation into $\lambdap$.
+
+\begin{thm}[Correctness of Input Sanitation for Translated Terms.]\label{thm:main}
+  If $\trden{e} = \iota$ and $\sctx \vdash e : \stringin{r}$ then $\iota \sreduces \tstr{s}$
+  for $s \in \lang{r}$.
+\end{thm}
+\begin{proof}
+  By theorem \ref{thm:tcorrect}, $\iota \sreduces \tstr{s}$ implies that $e \sreduces \strin{s}$.
+  By theorem \ref{thm:scorrect}, the above property together with the assumption that $e$ is well-typed implies that $s \in \lang{r}$.
+\end{proof}
 
@@
 
-\subsection{Definition of $\lambdas$}
-
-The $\lambdas$ system extends the simply-typed lambda calculus with regular
-expression types
-
-%\newpage
-
 \begin{figure}[t]\label{fig:lambdap}
@@
 \newpage
-
-\subsection{Definition of $\lambdap$}
 
@@
 
-The system $\lambdap$ is a straight-forward extension of a simply typed lambda calculus
-with a string type and a regular expression tpye. We include two operations which are available in the regular expression
-library of any modern programming language. The check operation ensures that an expression
-recognizes a string, and the replace operation is string replacement.
-
-\subsection{Definition of Translation}
-
 \begin{figure}[h]\label{fig:trans}
@@
 
-The translation from $\lambdas$ to $\lambdap$ is defined in figure 5.
-The coercion cases are most interesting. If the safety of coercion in manifest in the
-types of the expressions, then no runtime check is inserted.
-If the safety of coercion is not manifest in the types, then a check is inserted.
-
-In practice, the type of a replacement rarely matches a specification.
-Therefore, it is convienant in an implementation to always insert the appropriate
-coercion, and then only raise type errors when an automatically inserted coercion
-actually requires the insertion of a runtime check. Alternatively, this policy may 
-be codified in the type system itself using subtyping \cite{fulton12}.
-
-\subsection{Properties of Regular Languages}
-
-Our type safety proof for language S replies on a relationship between
-string substitution and language substitution given in lemma \ref{thm:substcorrespondence}.
-We also rely upon several other properties of regular languages.
-Throughout this section, we fix an alphabet $\Sigma$ over which strings $s$ and
-regular expressions $r$ are defined. throughout the paper, $\lang{r}$ refers to the
-language recognized by the regular expression $r$. This distinction between the regular expression
-and its language -- typically elided in the literature -- makes our definition and
-proofs about systems S and P more readable.
-
-\begin{lem}[Properties of Regular Languages and Expressions.] \label{thm:regexprops}
-The following are properties of regular expressions which are necessary for our proofs:
-If $s_1 \in \lang{r_1}$ and $s_2 \in \lang{r_2}$ then $s_1s_2 \in \lang{r_1r_2}$.
-For all strings $s$ and regular expressions $r$, either $s \in \lang{r}$ or $s \not \in \lang{r}$.
-Regular languages are closed under difference, right quotient, reversal, and string homomorphism.
-\end{lem}
-
-If any of these properties are unfamiliar, the reader may refer to a standard text on the subject \cite{cinderella}.
-
-\begin{defn}[$\tt{subst}$]
-The relation $\lsubst{r}{s_1}{s_2} = s$ produces a string $s$ in which all substrings of $s_1$ matching $r$ are replaced with $s_2$.
-\end{defn}
-
-\begin{defn}[$\tt{lreplace}$]
-  The relation $\lreplace{r}{r_1}{r_2} = r'$ relates $r, r_1,$ and $r_2$ to
-  a language $r'$ containing all strings of $r_1$ except that any substring $s_{pre} s s_{post} \in \lang{r_1}$ where $s \in \lang{r}$
-  is replaced by the set of strings $s_{pre} s_2 s_{post}$ for all $s_2 \in \lang{r_2}$ (the prefix and postfix positions may be empty).
-\end{defn}
-
-\begin{prop}[Closure.] \label{thm:total}
-  If $\lang{r}, \lang{r_1}$ and $\lang{r_2}$ are regular expressions, then $\lang{\lreplace{r}{r_1}{r_2}}$ is also a regular language.
-\end{prop}
-\begin{proof}
-The theorem follows from closure under difference, right quotient, reversal and string homomorphism.
-\end{proof}
-
-\begin{prop}[Substitution Correspondence.] \label{thm:substcorrespondence}
-  If $s_1 \in \lang{r_1}$ and $s_2 \in \lang{r_2}$ then $\lsubst{r}{s_1}{s_2} \in \lang{\lreplace{r}{s_1}{s_2}}$.
-\end{prop}
-\begin{proof}
-The theorem follows from the definitions of subst and lreplace; note that language substitutions over-approximate string substitutions.
-\end{proof}
-
-%\begin{description}
-%\item[case $r=\alpha$.]
-%If $s_1 = \alpha$ then $\alpha \in \lang{r_1}$ by assumption.
-%Therefore, $\lsubst{r}{s_1}{s_2} = \lsubst{\alpha}{\alpha}{s_2} = s_2$ and
-%$\lreplace{r}{r_1}{r_2} = \lreplace{\alpha}{r_1}{r_2}$. 
-%Since $s_1 = \alpha$ and $\alpha \in \lang{r_1}$, $r_1 \iso \alpha | r_1'$ for some $r_1'$.
-%Therefore, $\lreplace{\alpha}{r_1}{r_2} \iso \lreplace{\alpha}{\alpha|r_1'}{r_2}$ by Lemma X(1).
-%Finally, $s_2 \in \lang{r_2}$ which implies $s_1 \in \lang{r_2|r'}$.
-%If $s_1 \not = \alpha$ the $\lsubst{r}{s_1}{s_2} = s_1$ and $\lreplace{\alpha}{r_1}{r_2} = r_1$.
-%
-%\item[case $r = a|b$].
-%Note that $\subst{a|b}{s_1}{s_2} = \lsubst{a}{\lsubst{b}{s_1}{s_2}}{s_2}$
-%and $\lsubst{b}{s_1}{s_2} \in \lreplace{b}{r_1}{r_2}$ by induction.
-%Therefore, $\lsubst{a|b}{s_1}{s_2} \in \lreplace{a}{\lreplace{b}{r_1}{r_2}}{r_2}$ by induction and the definition of lreplace.
-%Finally, applying definitions once more, $\lsubst{a|b}{s_1}{s_2} \in \lreplace{a|b}{r_1}{r_2}$.
-%
-%\item[case $r = ab$].
-%By a similar argument to the disjunctive case.
-%
-%\item[case $r = a*$].
-%By considering the once unwinding of $a*$, noting that $s_1$ and $s_2$ are finite.
-%\end{description}
-
-% NRF The coercion lemma is no longer necessary, because it is a property of
-% "strip"-style sanitation rather than the more general case of replacement 
-% treated in this paper. However, we should mention this difference between my
-% undergraduate thesis and this paper in the related work section if we are
-% going tin clude a discussio of unpublished or quasi-published (e.g. src)
-% results.
-%\begin{lem}{Coercion Lemma.} \label{thm:coercionalemma}
-%Suppose that $R$ and $L$ are regular expressions, and that $s \in R$ is a finite string.  
-%Let $s' = \coerce(R,L,s)$ with all maximal substrings recognized by $L$ replaced with $\epsilon$.
-Then $s'$ is recognized by $(R \backslash L) + \emptyset$ and the construction of $R \backslash L$ is decidable.
-%\end{lem}
-%\begin{proof}
-%Let $F,G$ be FAs corresponding to $R$ and $L$, and let $G'$ be $G$ with its final states inverted (so that $G'$ is the complement of $L$).
-Define an FA $H$ as a DFA corresponding to the NFA found by combining $F$ and $G'$ such that $H$ accepts only if $R$ and $L'$ accept or if $s$ is empty (this construction may result in an exponential blowup in state size.)  Clearly, $H$ corresponds to $R \backslash L + \emptyset$.
-Thus, the construction of $R \backslash L + \emptyset$ is decidable.
-%
-%If $R \subset L$, $s' = \emptyset$.
-If $L \subset R$, either $s' = \emptyset$, or $s' \in R$ and $s' \not \in L$.
-If $R$ and $L$ are not subsets of one another, then it may be the case that $L$ recognizes part of $R$.
-Consider $L$ as the union of two languages, one which is a subset of $R$ and one which is disjoint.
-The subset language is considered above and the disjoint language is inconsequential.
-%\end{proof}
-
-% Why is this necessary?
-%\begin{lem}
-%  For all $r$, if $s_1 \in r_1$ and $s_2 \in r_2$ then $\subst{r}{s_1}{s_2} \subseteq \lsubst{r}{r_1}{r_2}$.
-%\end{lem}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-% BEGIN PROOFS ABOUT S AND P
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-\subsection{Safety of the Source and Target Languages}
-
-In this section, we establish type safety for the source ($\lambdas$) and
-target ($\lambdap$) languages. In addition to type safety, we also prove
-a stronger correctness property for $\lambdas$.
-
-Our first two theorems establish that our rules preserve the well-formedness of
-regular expressions. The standard lemmas required for safety of the simply
-typed lambda claculus are also required; proofs for these do not differ substntially
-from \cite{pfpl}.
-
-\begin{lem}
-  If $\sctx \vdash e : \stringin{r}$ then $r$ is a well-formed regular expression.
-\end{lem}
-\begin{proof}
-  The only non-trivial case is S-T-Replace, which follows from lemma \ref{thm:total}. 
-\end{proof}
-\begin{lem}
-If $\tctx \vdash \iota : \regex$ then $\iota \treduces \rx{r}$ such that $r$ is a well-formed regular expression. 
-\end{lem}
-
-We now prove safety for the string fragment of the source and target languages.
-The easiest property is type safety of $\lambdap$, which follows almost directly from
-type safety for the simply typed lambda calculus.
-
-\begin{thm} 
-  Let $\iota$ be a term in the target language. If $\tctx \vdash \iota : \tau$ 
-  then $\iota \sreduces \iota'$ and $\tctx \vdash \iota' : \tau$.
-\end{thm}
-
-Safety for the string fragment of $\lambdas$ is more involved because it involves validating that
-the type system's definition is justified by our theorems about regular languages.
-We state the theorem for the string fragment only to demontrate this point;
-full type safety for $\lambdas$ follows easily from safety for the string fagment.
-Note that type safety does \emph{not} guarantee that strings of a type are in the
-correct language. We isolate this property so that we may reason about its preservation
-under translation to $\lambdap$, which does not have regular expression types.
-
-\begin{thm}[Type Safety for the String Fragment of $\lambdas$.] \label{thm:scorrect}
-  If $\sctx \vdash e : \sigma$ 
-  then $e \sreduces e'$ and $\sctx \vdash e' : \stringin{r}$.
-\end{thm}
-\begin{proof}
-By induction on the typing relation.
-The S-T-Concat case requires Lemma \ref{thm:regexprops} 
-and the S-T-Replace case appeals to Lemma \ref{thm:substcorrespondence}.
-\end{proof}
-
-\begin{thm}[Canonical Forms for String Fragment of $\lambdas$.]\label{thm:cfs}
-  If $\sctx \vdash e : \stringin{r}$ 
-  then $e \sreduces \rstr{s}$ 
-\end{thm}
-
-In addition to safety, $\lambdas$ requires a correctness result ensuring that well-typed terms of regular string type are in the language associated with their type.
-
-\begin{thm}[Correctness of Input Sanitation for $\lambdas$.]\label{thm:scorrect}
-  If  $\sctx \vdash e : \stringin{r}$ and $e \sreduces \rstr{s}$ then $s \in \lang{r}$.
-\end{thm}
-\begin{proof}
-  Follows directly from type safety, canonical forms for $\lambdas$, and inversion of the typing relation for $\lambdas$.
-\end{proof}
-
-\subsection{Translation Correctness}
-
-The main theorem of this paper is Theorem \ref{thm:main},
-which establishes that Theorem \ref{thm:scorrect} is preserved under translation
-into the target language $\lambdap$.
-
-Establishing this result requires an additional theorem establishing a relationship between canonical forms for the string fragments of $\lambdas$ and $\lambdap$.
-
-\begin{thm}[Translation Correctness.] \label{thm:tcorrect}
-If $\sctx \vdash e : \stringin{r}$ then there exists an $\iota$ such that $\trden{e} = \iota$, 
-$\iota \treduces \tstr{s}$, and $e \sreduces \strin{s}$.
-\end{thm}
-
-\begin{proof}
-The proof proceeds by induction on the typing relation for $e$ and an appropriate
-choice of $\iota$; in each case, the choice is obvious and the rest of the proof
-proceeds by our type safety theorems as well as an appeal to the induction hypothesis.
-\end{proof}
-
-Finally, our main theorem establishes that input sanitation correctness of $\lambdas$ is preserved under the translation into $\lambdap$.
-
-\begin{thm}[Correctness of Input Sanitation for Translated Terms.]\label{thm:main}
-  If $\trden{e} = \iota$ and $\sctx \vdash e : \stringin{r}$ then $\iota \sreduces \tstr{s}$
-  for $s \in \lang{r}$.
-\end{thm}
-\begin{proof}
-  By theorem \ref{thm:tcorrect}, $\iota \sreduces \tstr{s}$ implies that $e \sreduces \strin{s}$.
-  By theorem \ref{thm:scorrect}, the above property together with the assumption that $e$ is well-typed implies that $s \in \lang{r}$.
-\end{proof}
-% PROOF OF TRANSLATION CORRECTNESS.
-% This proof is slightly out of say. It's SUFFICIENT, but is actually more
-% than sufficient because we prove \ref{scorrect} along with translation
-% correctness....
-%\begin{description}
-%\item[S-T-String-I:] 
-%Let $S = \rstr{s}$ and suppose $\sctx \vdash \rstr{s} : \stringin{r}$.
-%Choose $T = \str{s}$ and note that $\trden{S} = P$ by Tr-String.
-%By P-E-String, $P \treduces \str{s}$ and by S-E-String $S \treduces \rstr{s}$.
-%Finally, by inversion of S-T-Stringin-I, $s \in \lang{r}$.
-%
-%\item[S-T-Concat:]
-%Let $S = \rconcat{S_1}{S_2}$ and suppose $\sctx \vdash S : \stringin{r_1 r_2}$.
-%By inversion, $\sctx \vdash S_1 : \stringin{r_1}$.
-%It follows by induction that there exists a $P_1$ such that $\trden{S_1} = P_1$.
-%By a similar argument for $S_2$ and $r_2$, there exists a $P_2$ such that $\trden{S_2} = P_2$.
-%Choose $P = \concat{P_1}{P_2}$.
-%
-%We first prove property (a).
-%Note that $S_1$ and $P_1$ do not result in errors and that $S_1$ is well typed.
-%Therefore, $S_1 \sreduces \rstr{s_1}$ and $P_1 \treduces \str{s_1}$ for some $s_1 \in \lang{r_1}$ by theorems 8 and 9 respectively. % TODO-nrf make sure this numbering is correct
-%Similarly, $S_2 \sreduces \rstr{s_2}$ and $P_2 \treduces \str{s_2}$ for some $s_2 \in \lang{r_2}$.
-%Therefore, $S \sreduces \rstr{s_1 s_2}$ by S-E-Concat and $\concat{P_1}{P_2} \treduces \str{s_1 s_2}$ by P-E-Concat.
-%Finally, $s_1s_2 \in \lang{r_1}{r_2}$ by \ref{thm:regexprops}.
-%
-%Consider property (b).
-%If $S_1 \err$ then $P_1 \err$ by induction, and it follows that $S \err$ and $P \err$ by respective error propagation rules.
-%Similarly, if $S_2 \err$ then $P_2 \err$ and it follow that $S \err$ and $P \err$ by induction and propagation.
-%
-%\item[S-T-Replace:]
-%Let $S = \rreplace{r}{S_1}{S_2}$ and suppose $\sctx \vdash S : \stringin{r'}$ for some $s$.
-%By inversion of S-T-Replace, , $\sctx S_1 : \stringin{r_1}$ and $\sctx : \stringin{r_2}$ 
-%such that $\lsubst{r}{r_1}{r_2} = r'$.
-%By induction, there exists some $P_1, P_2$ such that $\trden{S_1} = P_1$, $\trden{S_2} = P_2$
-%and either (a) or (b) holds.
-%
-%If (a) holds then $S_1 \sreduces \rstr{s_1}$ and $P_1 \treduces \str{s_1}$ for some $s_1 \in \lang{r_1}$,
-%and similarly for $S_2, P_2$ and some $s_2 \in \lang{r_2}$.
-%Therefore, by S-E-Replace, $S \sreduces \rstr{s}$ for some $s = \lreplace{r}{s_1}{s_2}$.
-%Choose $P = \preplace{r}{s_1}{s_2}$.
-%By a similar argument and P-E-Replace, $P \treduces \str{s}$ for some $s = \lreplace{r}{s_1}{s_2}$.
-%What remains to be shown is $\lreplace{r}{s_1}{s_2} \in \lang{\lsubst{r}{r_1}{r_2}}$,
-%which follows from Leamm D since $s_1 \in r_1$ and $s_2 \in r_2$.
-%
-%If (b) holds for $S_1$ and $P_1$, then $S \err$ and $P \err$ by propagation rules.
-%Similarly, if (b) hols for $S_2$ and $P_2$ then $S \err$ and $P \err$ by propagation rules.
-%
-%\item[S-T-Coerce:]
-%Let $S = \rcoerce{r}{S'}$ and suppose $\sctx \vdash \rcoerce{r}{S} : \stringin{r}$.
-%By inverstion $\Psi \vdash S' : \stringin{r'}$ for an arbitrary $r'$.
-%By induction there exists a $P'$ such that $\trden{S'} = P'$ and either (a) or (b)
-%holds for $S'$ and $P'$.
-%
-%If (a) holds then $S' \sreduces \rstr{s'}$ and $P' \treduces \str{s'}$ for some $s' \in \lang{r'}$.
-%Note that either $s' \in \lang{r}$ or $s' \not \in \lang{r}$ by property 2 of \ref{thm:regexprops}.
-%Suppose $s' \in \lang{r}$. Then $\rcoerce{r}{S} \sreduces \rstr{S'}$ by S-E-Coerce.
-%Choose $P = \check{\rx{r}}{P'}$ and note that $P \treduces \str{s'}$ by P-E-Coerce.
-%Now suppose $s' \not \in \lang{r}$. Then $S \err$ and $P \err$ by P-E-Check-Err
-%and S-E-Coerce-Err.
-%
-%Finally, if (b) holds then $S \err$ and $P \err$ by propagation.
-%
-%\end{description}
-\section{Implementation in {Atlang}}\label{atlang}
 %\begin{figure}

### 2014-08-19 sanitation-psp14/sanitation-psp14-short.tex
Fixed layout of the paper.

@@
 \newcommand{\lreplace}[3]{{\sf lreplace}(#1; #2; #3)}
+
+\newcommand{\sctx}{\Psi} % Context for external typing
+\newcommand{\tctx}{\Theta} % Context for internal typing
+\newcommand{\ereduces}{\Downarrow}
 
@@
 \section{Regular String Types, Minimally}\label{calculus}
+
+In this section, we define a minimal typed lambda calculus with regular string types called $\lambda_{RS}$.
+The syntax of $\lambda_{RS}$ is specified in Figure \ref{fig:lamrs}, its static semantics in Figure \ref{fig:etyping} and its evaluation semantics in Figure \ref{fig:eeval}.
+This will serve as the source language for our translation to a calculus with only standard strings and regular expressions, defined in Sec.
+\ref{translation}.
+
+There are two type constructors in $\lambdas$, $\rightarrow$ and $\textsf{stringin}$.
+Arrow types classify functions, which are introduced via lambda abstraction, $\lambda x.e$, and can be applied, written $e(e)$, in the usual way \cite{pfpl}.
+Regular string types are of the form $\stringin{r}$, where $r$ is a regular expression.
+Values of such regular string types take the form $\strin{s}$, where $s$ is a string (i.e.
+$s \in \Sigma^\star$, defined in the usual way).
+The rule \textsc{S-T-Stringin-I} ensures that $s \in \lang{r}$.
+The remaining operations on terms of type $\stringin{r}$ preserve this property.
+%The remaining operators operate on regular strings.
+
+An outline for this section follows:
+\begin{itemize}
+\item In section \ref{sec:rs}, we describe $\lambdas$. We also prove provide proof outlines for type soundness and a correctness theorem.
+\item In section \ref{sec:p}, we present a simple target language with a minimal regular expression library.
+\item Section \ref{sec:tr} describes the translation from $\lambdas$ to $\lambdap$ and ensures the correcness result for \ref{sec:rs} is preserved under translation. 
+\end{itemize}
+
+\subsection{The Language of Constrained Strings}\label{sec:rs}
+
+The system $\lambdas$ provides many familiar operations on strings.
+However, unlike most languages, the type system of $\lambdas$ tracks the effect of these operations by relating computations on strings to corresponding computations on regular languages.
+This section describes our approach, culminating in a Security Theorem.
+The main technical result of this paper will show that this Security Theorem is preserved when $\lambdas$ is translated to a more traditional language.
+Throughout this section, we refer to \textsc{Inference Rules} defined in the figures contained in this section.
+\todo{remove tranditional, most and improve description}
+
+In this section, we define $\lambdas$, describe each operation, and finally prove both type safety and a correctness result for the language.
+The grammar for $\lambdas$ is defined in Figure \ref{fig:glambdas}. Typing rules \textsc{S-T-} are defined in Figure \ref{fig:slambdas} and a big-step semantics
+is defined in Figure \ref{fig:dlambdas}
+\footnote{For the reader's convienance, a single sheet containing figures \ref{fig:glambdas}-\ref{fig:tr} is available at http://nfulton.org/strings/printout.pdf \todo{post this}}.
+
 \renewcommand{\grammarlabel}[2]{#1\hfill#2}
-
-\newcommand{\sctx}{\Psi} % Context for external typing
-\newcommand{\tctx}{\Theta} % Context for internal typing
-
-\newcommand{\ereduces}{\Downarrow}
-
-\begin{figure}[t]
+\begin{figure}[h]
 \begin{grammar}
@@
       $\rcheck{r}{e}{x.e}{e}$
-\caption{Syntax of $\lambda_{RS}$.}\label{fig:lamrs}
+\caption{Syntax of $\lambda_{RS}$.}
+\label{fig:glambdas}
 \end{grammar}
@@
 
-In this section, we define a minimal typed lambda calculus with regular string types called $\lambda_{RS}$.
-The syntax of $\lambda_{RS}$ is specified in Figure \ref{fig:lamrs}, its static semantics in Figure \ref{fig:etyping} and its evaluation semantics in Figure \ref{fig:eeval}.
-This will serve as the source language for our translation to a calculus with only standard strings and regular expressions, defined in Sec.
-\ref{translation}.
-
-There are two type constructors in $\lambdas$, $\rightarrow$ and $\textsf{stringin}$.
-Arrow types classify functions, which are introduced via lambda abstraction, $\lambda x.e$, and can be applied, written $e(e)$, in the usual way \cite{pfpl}.
-Regular string types are of the form $\stringin{r}$, where $r$ is a regular expression.
-Values of such regular string types take the form $\strin{s}$, where $s$ is a string (i.e.
-$s \in \Sigma^\star$, defined in the usual way).
-The rule \textsc{S-T-Stringin-I} ensures that $s \in \lang{r}$.
-The remaining operations on terms of type $\stringin{r}$ preserve this property.
-%The remaining operators operate on regular strings.
-
-An outline for this section follows:
-\begin{itemize}
-\item In section \ref{sec:rs}, we describe $\lambdas$. We also prove provide proof outlines for type soundness and a correctness theorem.
-\item In section \ref{sec:p}, we present a simple target language with a minimal regular expression library.
-\item Section \ref{sec:tr} describes the translation from $\lambdas$ to $\lambdap$ and ensures the correcness result for \ref{sec:rs} is preserved under translation. 
-\end{itemize}
-
-\subsection{The Language of Constrained Strings}\label{sec:rs}
-
-The system $\lambdas$ provides many familiar operations on strings.
-However, unlike most languages, the type system of $\lambdas$ tracks the effect of these operations by relating computations on strings to corresponding computations on regular languages.
-This section describes our approach, culminating in a Security Theorem.
-The main technical result of this paper will show that this Security Theorem is preserved when $\lambdas$ is translated to a more traditional language.
-Throughout this section, we refer to \textsc{Inference Rules} defined in the figures contained in this section.
-\todo{remove tranditional, most and improve description}
+\begin{figure}[h]
+\small
+$\fbox{\inferrule{}{\sctx \vdash e : \sigma}}$
+~~~~$\sctx ::= \emptyset \pipe \sctx, x : \sigma$
+\begin{mathpar}
+\inferrule[S-T-Var]
+{ x:\sigma \in \sctx }
+{ \sctx \vdash x:\sigma}
+
+\inferrule[S-T-Abs]
+{\sctx, x : \sigma_1 \vdash e : \sigma_2}
+{\sctx \vdash \lambda x.e : \sigma_1 \rightarrow \sigma_2}
+
+\inferrule[S-T-App]
+{\sctx \vdash e_1 : \sigma_2 \rightarrow \sigma \\ \sctx \vdash e_2 : \sigma_2}
+{\sctx \vdash e_1(e_2) : \sigma}
+
+\inferrule[S-T-Stringin-I]
+{s \in \lang{r}}
+{\sctx \vdash \strin{s} : \stringin{r}}
+
+\inferrule[S-T-Concat]
+{\sctx \vdash e_1 : \stringin{r_1} \\ \sctx \vdash e_2 : \stringin{r_2}}
+{\sctx \vdash \rsconcat{e_1}{e_2} : \stringin{r_1 \cdot r_2}}
+
+\inferrule[S-T-Case]
+{ \sctx \vdash e_1 : \stringin{r} \\
+  \sctx \vdash e_2 : \sigma \\
+  \sctx, x : \stringin{\lhead{r}}, y : \stringin{\ltail{r}} \vdash e_3 : \sigma
+}
+{
+  \sctx \vdash \strcase{e_1}{e_2}{x.y.e_3} : \sigma
+}
+
+\inferrule[S-T-Replace]
+{\sctx \vdash e_1 : \stringin{r_1} \\ \sctx \vdash e_2 : \stringin{r_2} \\\\ \lreplace{r}{r_1}{r_2} = r'}
+{\sctx \vdash \sisubst{r}{e_1}{e_2} : \stringin{r'}}
+
+\inferrule[S-T-SafeCoerce]
+{\sctx \vdash e : \stringin{r'} \\ \lang{r'} \subseteq \lang{r}}
+{\sctx \vdash \coerce{r}{e} : \stringin{r}}
+
+\inferrule[S-T-Check]
+{\sctx \vdash e_0 : \stringin{r_0} \\ \sctx, x:\stringin{r} \vdash e_1 : \sigma \\ \sctx \vdash e_2 : \sigma}
+{\sctx \vdash \rcheck{r}{e_0}{x.e_1}{e_2} : \sigma}
+
+\end{mathpar}
+
+\caption{Typing rules for $\lambdas$. 
+The typing context $\sctx$ is standard.}
+\label{fig:slambdas}
+\end{figure}
+
+\begin{figure}
+  \small
+$\fbox{\inferrule{}{e \sreduces e}}$
+\begin{mathpar}
+\inferrule[S-E-Abs]                                                             
+{ \ }                                                                           
+{\lambda x.e \sreduces \lambda x.e}
+
+\inferrule[S-E-App]
+{ e_1 \sreduces \lambda x . e_3 \\ e_2 \sreduces e_2' \\ [ e_2' / x ] e_3 \sreduces v }
+{ e_1(e_2) \sreduces v }
+
+\inferrule[S-E-RStr]
+{ \ }
+{\strin{s} \sreduces \strin{s}}
+
+\inferrule[S-E-Concat]
+{e_1 \sreduces \strin{s_1} \\ e_2 \sreduces \strin{s_2}} 
+{\rsconcat{e_1}{e_2} \sreduces \strin{s_1 s_2}} % ???
+
+\inferrule[S-E-Case-$\epsilon$]
+{
+  e_1 \sreduces \strin{\epsilon} \\
+  e_2 \sreduces v_2
+}
+{
+  \strcase{e_1}{e_2}{e_3} \sreduces v_2
+}
+
+\inferrule[S-E-Case-Concat]
+{
+  e_1 \sreduces \strin{p s} \\
+  e_3 \sreduces x.y.e_4 \\
+  [p/x][s/y]e_4 \sreduces v
+}
+{
+  \strcase{e_1}{e_2}{e_3} \sreduces v
+}
+
+\inferrule[S-E-Replace]
+{e_1 \sreduces \strin{e_1} \\ e_2 \sreduces \strin{s_2} \\ \lsubst{r}{s_1}{s_2} = s} 
+{\sisubst{r}{e_1}{e_2} \sreduces \sistr{s}}
+
+\inferrule[S-E-SafeCoerce]
+{e \sreduces \strin{s}}
+{\coerce{r}{e} \sreduces \strin{s}}
+
+\inferrule[S-E-Check-Ok]
+{e \sreduces \strin{s} \\ s \in \lang{r} \\ [\strin{s} / x]e_1 \sreduces v}
+{
+  \rcheck{r}{e}{x.e_1}{e_2} \sreduces v
+}
+
+\inferrule[S-E-Check-NotOk]
+{
+  e \sreduces \strin{s} \\ s \not \in \lang{r}
+}
+{
+  \rcheck{r}{e}{x.e_1}{e_2} \sreduces e_2
+}
+
+\end{mathpar}
+\caption{Big step semantics for $\lambdas$}.
+\label{fig:dlambdas}
+\end{figure}
 
@@
 
-Type safety for $\lambdap$ is straight-forward, but is necessary in order to establish the 
-correctness of our translation.
-
-\begin{thm} 
-  Let $\iota$ be a term in the target language. If $\tctx \vdash \iota : \tau$ 
-  then $\iota \sreduces \iota'$ and $\tctx \vdash \iota' : \tau$.
-\end{thm}
-
-\subsection{Translation from $\lambdas$ to $\lambdap$}\label{sec:tr}
-
-The translation from $\lambdas$ to $\lambdap$ is defined in figure 5.
-The coercion cases are most interesting. If the safety of coercion in manifest in the
-types of the expressions, then no runtime check is inserted.
-If the safety of coercion is not manifest in the types, then a check is inserted.
-
-In practice, the type of a replacement rarely matches a specification.
-Therefore, it is convienant in an implementation to always insert the appropriate
-safe coercion. This way, violations of the sanitation protocol manifest as type errors,
-but the definition of an adequate replacement is not arduous.
-Alternatively, this policy may 
-be codified in the type system itself using subtyping \cite{fulton12}.
-
-This section ultimately establishes that the security theorem for $\lambdas$ is preserved under compilation.
-
-\begin{thm}[Translation Correctness]\label{thm:trcorrect}
-  If $\tctx \vdash e : \stringin{r}$ then there exists an $\iota$ such that $\trden{e} = \iota$, $\iota \sreduces \tstr{s}$,
-  and $e \treduces \rstr{s}$.
-\end{thm}
-\begin{proof}
-The proof proceeds by induction on the typing relation for $e$ and an appropraite choich of $\iota$; in each case,
-the choice is the syntactic form in $\lambdap$ corresponding to the form under consideration (e.g. choose {\sf preplace} when considering {\sf sreplace}).
-After the correct choice, the proof proceeds by our type safety theorems and an appeal to the induction hypothesis.
-\end{proof}
-
-Finally, our main result establishes that correctness of $\lambdas$ is preserved under the translation into $\lambdap$.
-
-\begin{thm}[Correctness of Input Sanitation for Translated Terms.]\label{thm:main}
-  If $\trden{e} = \iota$ and $\sctx \vdash e : \stringin{r}$ then $\iota \sreduces \tstr{s}$
-  for $s \in \lang{r}$.
-\end{thm}
-\begin{proof}
-  By theorem \ref{thm:tcorrect}, $\iota \sreduces \tstr{s}$ implies that $e \sreduces \strin{s}$.
-  By theorem \ref{thm:scorrect}, the above property together with the assumption that $e$ is well-typed implies that $s \in \lang{r}$.
-\end{proof}
-
-\begin{figure}\label{fig:lambdas}
-\small
-$\fbox{\inferrule{}{\sctx \vdash e : \sigma}}$
-~~~~$\sctx ::= \emptyset \pipe \sctx, x : \sigma$
-\begin{mathpar}
-\inferrule[S-T-Var]
-{ x:\sigma \in \sctx }
-{ \sctx \vdash x:\sigma}
-
-\inferrule[S-T-Abs]
-{\sctx, x : \sigma_1 \vdash e : \sigma_2}
-{\sctx \vdash \lambda x.e : \sigma_1 \rightarrow \sigma_2}
-
-\inferrule[S-T-App]
-{\sctx \vdash e_1 : \sigma_2 \rightarrow \sigma \\ \sctx \vdash e_2 : \sigma_2}
-{\sctx \vdash e_1(e_2) : \sigma}
-
-\inferrule[S-T-Stringin-I]
-{s \in \lang{r}}
-{\sctx \vdash \strin{s} : \stringin{r}}
-
-\inferrule[S-T-Concat]
-{\sctx \vdash e_1 : \stringin{r_1} \\ \sctx \vdash e_2 : \stringin{r_2}}
-{\sctx \vdash \rsconcat{e_1}{e_2} : \stringin{r_1 \cdot r_2}}
-
-\inferrule[S-T-Case]
-{ \sctx \vdash e_1 : \stringin{r} \\
-  \sctx \vdash e_2 : \sigma \\
-  \sctx, x : \stringin{\lhead{r}}, y : \stringin{\ltail{r}} \vdash e_3 : \sigma
-}
-{
-  \sctx \vdash \strcase{e_1}{e_2}{x.y.e_3} : \sigma
-}
-
-\inferrule[S-T-Replace]
-{\sctx \vdash e_1 : \stringin{r_1} \\ \sctx \vdash e_2 : \stringin{r_2} \\\\ \lreplace{r}{r_1}{r_2} = r'}
-{\sctx \vdash \sisubst{r}{e_1}{e_2} : \stringin{r'}}
-
-\inferrule[S-T-SafeCoerce]
-{\sctx \vdash e : \stringin{r'} \\ \lang{r'} \subseteq \lang{r}}
-{\sctx \vdash \coerce{r}{e} : \stringin{r}}
-
-\inferrule[S-T-Check]
-{\sctx \vdash e_0 : \stringin{r_0} \\ \sctx, x:\stringin{r} \vdash e_1 : \sigma \\ \sctx \vdash e_2 : \sigma}
-{\sctx \vdash \rcheck{r}{e_0}{x.e_1}{e_2} : \sigma}
-
-\end{mathpar}
-
-\caption{Typing rules for $\lambdas$. 
-The typing context $\sctx$ is standard.}
-\label{fig:etyping}
-\end{figure}
-
-\begin{figure}
-\small
-$\fbox{\inferrule{}{e \sreduces e}}$
-\begin{mathpar}
-\inferrule[S-E-Abs]                                                             
-{ \ }                                                                           
-{\lambda x.e \sreduces \lambda x.e}
-
-\inferrule[S-E-App]
-{ e_1 \sreduces \lambda x . e_3 \\ e_2 \sreduces e_2' \\ [ e_2' / x ] e_3 \sreduces v }
-{ e_1(e_2) \sreduces v }
-
-\inferrule[S-E-RStr]
-{ \ }
-{\strin{s} \sreduces \strin{s}}
-
-\inferrule[S-E-Concat]
-{e_1 \sreduces \strin{s_1} \\ e_2 \sreduces \strin{s_2}} 
-{\rsconcat{e_1}{e_2} \sreduces \strin{s_1 s_2}} % ???
-
-\inferrule[S-E-Case-$\epsilon$]
-{
-  e_1 \sreduces \strin{\epsilon} \\
-  e_2 \sreduces v_2
-}
-{
-  \strcase{e_1}{e_2}{e_3} \sreduces v_2
-}
-
-\inferrule[S-E-Case-Concat]
-{
-  e_1 \sreduces \strin{p s} \\
-  e_3 \sreduces x.y.e_4 \\
-  [p/x][s/y]e_4 \sreduces v
-}
-{
-  \strcase{e_1}{e_2}{e_3} \sreduces v
-}
-
-\inferrule[S-E-Replace]
-{e_1 \sreduces \strin{e_1} \\ e_2 \sreduces \strin{s_2} \\ \lsubst{r}{s_1}{s_2} = s} 
-{\sisubst{r}{e_1}{e_2} \sreduces \sistr{s}}
-
-\inferrule[S-E-SafeCoerce]
-{e \sreduces \strin{s}}
-{\coerce{r}{e} \sreduces \strin{s}}
-
-\inferrule[S-E-Check-Ok]
-{e \sreduces \strin{s} \\ s \in \lang{r} \\ [\strin{s} / x]e_1 \sreduces v}
-{
-  \rcheck{r}{e}{x.e_1}{e_2} \sreduces v
-}
-
-\inferrule[S-E-Check-NotOk]
-{
-  e \sreduces \strin{s} \\ s \not \in \lang{r}
-}
-{
-  \rcheck{r}{e}{x.e_1}{e_2} \sreduces e_2
-}
-
-\end{mathpar}
-\caption{Big step semantics for $\lambdas$}.
-\label{fig:eeval}
-\end{figure}
-
-%
-% Begin typing/translation relation
-%
-
-% Begin Grammar 
-% Removes the < > from grammar productions.
+The grammar of $\lambdap$ is defined in Figure~\ref{fig:lcsSyntax}.
+The typing rules \textsc{P-T-} are defined in Figure~\ref{fig:slambdap}
+and a big-step semantics is defined by the rules \textsc{P-E-} in Figure~\ref{fig:dlambdap}.
+
 \renewcommand{\grammarlabel}[2]{#1\hfill#2}
-
-% \begin{figure}[t]
-% \begin{grammar}
-% <r> ::= $\epsilon$ | $.$ | $a$ | $r \cdot r$ | $r + r$ | $r*$ \hfill $a \in \Sigma$
-
-% \caption{Regular expressions over the alphabet $\Sigma$.}
-% \label{fig:regex}
-% \end{grammar}
-% \end{figure}
-
-% \begin{figure}[h]
-% \begin{grammar}
-
-% <$\psi$> ::=	$\psi \rightarrow \psi$			\hfill	source types					\alt
-% $\stringin{r}$
-
-% <S> ::= 
-%       $\lambda x . e$ \hfill source terms \alt
-%       $e e$ \alt
-%       $\strin{s}$ \hfill $s \in \Sigma^{*}$ \alt
-%       $\rsconcat{S}{S}$ \alt
-%       $\sisubst{r}{S}{S}$ \alt
-%       $\coerce{r}{S}$
-% \caption{Syntax for the string sanitation fragment of our source language, $\lambdas$.}
-% \end{grammar}
-% \end{figure}
-
-\begin{figure}[t]\label{fig:lambdap}
-\small
-$\fbox{\inferrule{}{\tctx \vdash \iota : \tau}}$
-~~~~$\tctx ::= \emptyset \pipe \tctx, x : \tau$
-
-\begin{mathpar}
-\inferrule[P-T-Var]
-{ x:\tau \in \tctx }
-{ \tctx \vdash x:\tau }
-
-\inferrule[P-T-Abs]
-{\tctx, x : \tau_1 \vdash \iota_2 : \tau_2}
-{\tctx \vdash \lambda x.\iota_2 : \tau_1 \rightarrow \tau_2}
-
-\inferrule[P-T-App]
-{\tctx \vdash \iota_1 : \tau_2 \rightarrow \tau \\ \tctx \vdash \iota_2 : \tau_2}
-{\tctx \vdash \iota_1(\iota_2) : \iota}
-
-\inferrule[P-T-String]
-{ \ }
-{\tctx \vdash \tstr{s} : \str}
-
-\inferrule[P-T-Regex]
-{ \ }
-{\tctx \vdash \rx{r} : \regex}
-
-\inferrule[P-T-Concat]
-{\tctx \vdash \iota_1 : \str \\ \tctx \vdash \iota_2 : \str}
-{\tctx \vdash \tconcat{\iota_1}{\iota_2} : \str}
-
-\inferrule[P-T-Case]
-{\tctx \vdash \iota_1 : \str \\ \tctx \vdash \iota_2 : \tau \\ \tctx, x:\str, y:\str \vdash \iota_3:\tau}
-{\tctx \vdash \pstrcase{\iota_1}{\iota_2}{\iota_3} : \tau}
-
-\inferrule[P-T-Replace]
-{\tctx \vdash \iota_1 : \regex \\ \tctx \vdash \iota_2 : \str \\ \tctx \vdash \iota_3 : \str }
-{\tctx \vdash \preplace{\iota_1}{\iota_2}{\iota_3} : \str}
-
-\inferrule[P-T-Check]
-{\tctx \vdash \iota_r : \regex \\ \tctx \vdash \iota_1 : \str \\ \tctx, x:\str \vdash \iota_2 : \sigma \\ \tctx \vdash \iota_3 : \sigma}
-{\tctx \vdash \tcheck{\iota_r}{\iota_1}{x . \iota_2}{\iota_3} : \sigma}
-\end{mathpar}
-\caption{Typing rules for $\lambdap$.
-The typing context $\tctx$ is standard.}
-\label{fig:ityping}
-\end{figure}
-
-\begin{figure}[t]
-\small
-$\fbox{\inferrule{}{\iota \treduces \iota}}$
-
-\begin{mathpar}
-\inferrule[P-E-Abs]
-{ \ }
-{\lambda x.e \sreduces \lambda x.e}
-
-\inferrule[P-E-App]
-{ \iota_1 \sreduces \lambda x . \iota_3 \\  \iota_2 \sreduces \iota_2' \\ [\iota_3 / x] \iota_2' \sreduces v}
-{ \iota_1(\iota_2) \sreduces v}
-
-\inferrule[P-E-Str]
-{ \ }
-{\tstr{s} \treduces \tstr{s}}
-
-\inferrule[P-E-Rx]
-{ \ }
-{\rx{r} \treduces \rx{r}}
-
-\inferrule[P-E-Concat]
-{\iota_1 \treduces \tstr{s_1} \\ \iota_2 \treduces \tstr{s_2}} 
-{\tconcat{\iota_1}{\iota_2} \treduces \tstr{s_1 s_2}} % ???
-
-\inferrule[P-E-Case-$\epsilon$]
-{
-  \iota_1 \treduces \tstr{\ } \\
-  \iota2 \treduces \iota_2
-}
-{
-  \pstrcase{\iota_1}{\iota_2}{\iota_3} \treduces v_2
-}
-
-\inferrule[P-E-Case-Concat]
-{
-  \iota_1 \treduces \tstr{ps} \\
-  x.y.\iota_3 \treduces \iota_4 \\
-  [p/x][s/y]\iota_4 \treduces v
-}
-{
-  \pstrcase{\iota_1}{\iota_2}{\iota_3} \treduces v
-}
-
-\inferrule[P-E-Replace]
-{\iota_1 \treduces \rx{r} \\ \iota_2 \treduces \tstr{s_2} \\ \iota_3 \treduces \tstr{s_3} \\ \lsubst{r}{s_2}{s_3} = s} 
-{\preplace{\iota_1}{\iota_2}{\iota_3} \treduces \tstr{s}}
-
-\inferrule[P-E-Check-OK]
-{\iota \treduces \tstr{s} \\ s \in \lang{r} \\ [\tstr{s} / x] \iota_1 \sreduces \iota_3}
-{\tcheck{\rx{r}}{\iota}{x.\iota_1}{\iota_2} \treduces \iota_3}
-
-\inferrule[P-E-Check-NotOK]
-{\iota \treduces \tstr{s} \\ s \not \in \lang{r}}
-{\tcheck{\rx{r}}{\iota}{x.\iota_1}{\iota_2} \treduces \iota_2}
-\end{mathpar}
-\caption{Big step semantics for of $\lambdap$}.
-\label{fig:ieval}
-\end{figure}
-
-\newpage
-\ 
-\newpage
-\ 
-\newpage
 
@@
 
-\begin{figure}[h]\label{fig:trans}
+\begin{figure}[h]\label{fig:lambdap}
+\small
+$\fbox{\inferrule{}{\tctx \vdash \iota : \tau}}$
+~~~~$\tctx ::= \emptyset \pipe \tctx, x : \tau$
+
+\begin{mathpar}
+\inferrule[P-T-Var]
+{ x:\tau \in \tctx }
+{ \tctx \vdash x:\tau }
+
+\inferrule[P-T-Abs]
+{\tctx, x : \tau_1 \vdash \iota_2 : \tau_2}
+{\tctx \vdash \lambda x.\iota_2 : \tau_1 \rightarrow \tau_2}
+
+\inferrule[P-T-App]
+{\tctx \vdash \iota_1 : \tau_2 \rightarrow \tau \\ \tctx \vdash \iota_2 : \tau_2}
+{\tctx \vdash \iota_1(\iota_2) : \iota}
+
+\inferrule[P-T-String]
+{ \ }
+{\tctx \vdash \tstr{s} : \str}
+
+\inferrule[P-T-Regex]
+{ \ }
+{\tctx \vdash \rx{r} : \regex}
+
+\inferrule[P-T-Concat]
+{\tctx \vdash \iota_1 : \str \\ \tctx \vdash \iota_2 : \str}
+{\tctx \vdash \tconcat{\iota_1}{\iota_2} : \str}
+
+\inferrule[P-T-Case]
+{\tctx \vdash \iota_1 : \str \\ \tctx \vdash \iota_2 : \tau \\ \tctx, x:\str, y:\str \vdash \iota_3:\tau}
+{\tctx \vdash \pstrcase{\iota_1}{\iota_2}{\iota_3} : \tau}
+
+\inferrule[P-T-Replace]
+{\tctx \vdash \iota_1 : \regex \\ \tctx \vdash \iota_2 : \str \\ \tctx \vdash \iota_3 : \str }
+{\tctx \vdash \preplace{\iota_1}{\iota_2}{\iota_3} : \str}
+
+\inferrule[P-T-Check]
+{\tctx \vdash \iota_r : \regex \\ \tctx \vdash \iota_1 : \str \\ \tctx, x:\str \vdash \iota_2 : \sigma \\ \tctx \vdash \iota_3 : \sigma}
+{\tctx \vdash \tcheck{\iota_r}{\iota_1}{x . \iota_2}{\iota_3} : \sigma}
+\end{mathpar}
+\caption{Typing rules for $\lambdap$.
+The typing context $\tctx$ is standard.}
+\label{fig:slambdap}
+\end{figure}
+
+\begin{figure}[t]
+\small
+$\fbox{\inferrule{}{\iota \treduces \iota}}$
+
+\begin{mathpar}
+\inferrule[P-E-Abs]
+{ \ }
+{\lambda x.e \sreduces \lambda x.e}
+
+\inferrule[P-E-App]
+{ \iota_1 \sreduces \lambda x . \iota_3 \\  \iota_2 \sreduces \iota_2' \\ [\iota_3 / x] \iota_2' \sreduces v}
+{ \iota_1(\iota_2) \sreduces v}
+
+\inferrule[P-E-Str]
+{ \ }
+{\tstr{s} \treduces \tstr{s}}
+
+\inferrule[P-E-Rx]
+{ \ }
+{\rx{r} \treduces \rx{r}}
+
+\inferrule[P-E-Concat]
+{\iota_1 \treduces \tstr{s_1} \\ \iota_2 \treduces \tstr{s_2}} 
+{\tconcat{\iota_1}{\iota_2} \treduces \tstr{s_1 s_2}} % ???
+
+\inferrule[P-E-Case-$\epsilon$]
+{
+  \iota_1 \treduces \tstr{\ } \\
+  \iota2 \treduces \iota_2
+}
+{
+  \pstrcase{\iota_1}{\iota_2}{\iota_3} \treduces v_2
+}
+
+\inferrule[P-E-Case-Concat]
+{
+  \iota_1 \treduces \tstr{ps} \\
+  x.y.\iota_3 \treduces \iota_4 \\
+  [p/x][s/y]\iota_4 \treduces v
+}
+{
+  \pstrcase{\iota_1}{\iota_2}{\iota_3} \treduces v
+}
+
+\inferrule[P-E-Replace]
+{\iota_1 \treduces \rx{r} \\ \iota_2 \treduces \tstr{s_2} \\ \iota_3 \treduces \tstr{s_3} \\ \lsubst{r}{s_2}{s_3} = s} 
+{\preplace{\iota_1}{\iota_2}{\iota_3} \treduces \tstr{s}}
+
+\inferrule[P-E-Check-OK]
+{\iota \treduces \tstr{s} \\ s \in \lang{r} \\ [\tstr{s} / x] \iota_1 \sreduces \iota_3}
+{\tcheck{\rx{r}}{\iota}{x.\iota_1}{\iota_2} \treduces \iota_3}
+
+\inferrule[P-E-Check-NotOK]
+{\iota \treduces \tstr{s} \\ s \not \in \lang{r}}
+{\tcheck{\rx{r}}{\iota}{x.\iota_1}{\iota_2} \treduces \iota_2}
+\end{mathpar}
+\caption{Big step semantics for of $\lambdap$}.
+\label{fig:dlambdap}
+\end{figure}
+
+\subsubsection{Safety}
+
+Type safety for $\lambdap$ is straight-forward, but is necessary in order to establish the 
+correctness of our translation.
+
+\begin{thm} 
+  Let $\iota$ be a term in the target language. If $\tctx \vdash \iota : \tau$ 
+  then $\iota \sreduces \iota'$ and $\tctx \vdash \iota' : \tau$.
+\end{thm}
+
+\subsection{Translation from $\lambdas$ to $\lambdap$}\label{sec:tr}
+
+The translation from $\lambdas$ to $\lambdap$ is defined in figure 5.
+The coercion cases are most interesting. If the safety of coercion in manifest in the
+types of the expressions, then no runtime check is inserted.
+If the safety of coercion is not manifest in the types, then a check is inserted.
+
+In practice, the type of a replacement rarely matches a specification.
+Therefore, it is convienant in an implementation to always insert the appropriate
+safe coercion. This way, violations of the sanitation protocol manifest as type errors,
+but the definition of an adequate replacement is not arduous.
+Alternatively, this policy may 
+be codified in the type system itself using subtyping \cite{fulton12}.
+
+The translation is defined by the rules \textsc{Tr-} in Figure~\ref{fig:tr}.
+This section ultimately establishes that the security theorem for $\lambdas$ is preserved under compilation.
+
+\begin{figure}[h]
 
@@
 The translation is type-directed.}
+\label{fig:tr}
 \end{figure}
+
+\begin{thm}[Translation Correctness]\label{thm:trcorrect}
+  If $\tctx \vdash e : \stringin{r}$ then there exists an $\iota$ such that $\trden{e} = \iota$, $\iota \sreduces \tstr{s}$,
+  and $e \treduces \rstr{s}$.
+\end{thm}
+\begin{proof}
+The proof proceeds by induction on the typing relation for $e$ and an appropraite choich of $\iota$; in each case,
+the choice is the syntactic form in $\lambdap$ corresponding to the form under consideration (e.g. choose {\sf preplace} when considering {\sf sreplace}).
+After the correct choice, the proof proceeds by our type safety theorems and an appeal to the induction hypothesis.
+\end{proof}
+
+\subsubsection{Preservation of Security}
+
+Finally, our main result establishes that correctness of $\lambdas$ is preserved under the translation into $\lambdap$.
+
+\begin{thm}[Correctness of Input Sanitation for Translated Terms.]\label{thm:main}
+  If $\trden{e} = \iota$ and $\sctx \vdash e : \stringin{r}$ then $\iota \sreduces \tstr{s}$
+  for $s \in \lang{r}$.
+\end{thm}
+\begin{proof}
+  By theorem \ref{thm:tcorrect}, $\iota \sreduces \tstr{s}$ implies that $e \sreduces \strin{s}$.
+  By theorem \ref{thm:scorrect}, the above property together with the assumption that $e$ is well-typed implies that $s \in \lang{r}$.
+\end{proof}
 
@@
 
-\subsection{An Example}\label{whatev}
+%\subsection{An Example}\label{whatev}
 
@@
 
-\label{fig:impl}
 \begin{figure}
@@
 \caption{Implementation of the \texttt{string\_in} type constructor in atlang.}
+\label{fig:impl}
+
 \label{fig:atimpl}\end{figure}

### 2014-08-19 sanitation-psp14/sanitation-psp14-short.tex
More layout work

@@
 
+\subsection{Example Usage}
+
 \begin{figure}[h]\begin{lstlisting}
@@
 
-Figure \ref{examplefig} demonstrates the use of string types in atlang.
+Figure \ref{fig:atexample} demonstrates the use of two type constructors, \verb|fn| and \verb|string_in|, both of which we have included in \verb|atlib|, the standard library for \verb|atlang|.
+The \verb|fn| type constructor can be used to annotate functions that should be statically checked by \verb|atlang|.
+The function \verb|sanitize| on lines 3-7 specifies one argument, \verb|s|, of type \lstinline{string_in[r'.*']}.\todo{fix sizes}
+
 The sanitize function takes an arbitrary string and returns a string without
@@
 
-The result of applying sanitize to input is appended to two functions which
-construct a safe query and safe output.
+The result of applying sanitize to input is appended in two functions which
+construct a safe query (avoiding command injection) and safe HTML output (avoiding XSS-style attacks).
 The arguments to the result and output construction functions constitute \emph{specifications}.
@@
 not contain any HTML tags, which is a conservative but effective policy for preventing
-XSS attacks.
+Cross-Site Scripting (XSS) attacks.
 
@@
 types.
+
 This subtyping relation is justified theoretically by the fact that language inclusion is
@@
 \label{fig:atimpl}\end{figure}
-
-\subsection{Example Usage}
-
-Figure \ref{fig:atexample} demonstrates the use of two type constructors, \verb|fn| and \verb|string_in|, both of which we have included in \verb|atlib|, the standard library for \verb|atlang|.
-The \verb|fn| type constructor can be used to annotate functions that should be statically checked by \verb|atlang|.
-The function \verb|sanitize| on lines 3-7 specifies one argument, \verb|s|, of type \lstinline{string_in[r'.*']}.\todo{fix sizes}
-
-The sanitize function takes an arbitrary string and returns a string without
-double quotes or left and right brackets. In this example, we use HTML
-escape sequences.
-
-The main function receives user input and passes this input to a 
-sanitize function, which replaces all double quotes and brackets with HTML
-escape sequences.
-
-The result of applying sanitize to input is appended to two functions which
-construct a safe query and safe output.
-The arguments to the result and output construction functions constitute \emph{specifications}.
-In the case of \textsf{results_query}, this specification ensures that user input is always interpreted
-as a string literal by the SQL server.
-In the case of \textsf{results_div}, this specification ensures that user input does 
-not contain any HTML tags, which is a conservative but effective policy for preventing
-XSS attacks.
-
-Note that ${\sf input}$ does not actually meet these specifications without additional machinery.
-The type of ${\sf input}$ is quite large
-and does not actually equal the specified domains of the query or output construction
-methods. This mismatch is common -- in fact, nearly universal. Therefore, 
-our implementation includes a simple subtyping relation between regular expression
-types. 
-This subtyping relation is justified theoretically by the fact that language inclusion is
-decidable; see \cite{fulton12} for a formal definition of the subtyping relation.
-Additionally, our extension remains composable because subtyping is defined on a type-by-type basis;
-see \cite{fulton13} for a discussion of subtyping in Atlang (referred to there as Ace).
-
-\subsection{Implementation of the Regular Expression Type}
 

### 2014-08-19 sanitation-psp14/sanitation-psp14-short.tex
Added defsheet.

@@
 is defined in Figure \ref{fig:dlambdas}
-\footnote{For the reader's convienance, a single sheet containing figures \ref{fig:glambdas}-\ref{fig:tr} is available at http://nfulton.org/strings/printout.pdf \todo{post this}}.
+\footnote{For the reader's convienance, a single sheet containing figures \ref{fig:glambdas}-\ref{fig:tr} is available at http://nfulton.org/strings/printout.pdf.}.
 

### 2014-08-19 sanitation-psp14/sanitation-psp14-short.tex
fugure work.

@@
 \newtheorem{rem}[thm]{Remark} 
+\renewcommand*{\proofname}{Proof Sketch}
 
@@
 For all strings $s$ and regular expressions $r$, either $s \in \lang{r}$ or $s \not \in \lang{r}$.
-Regular languages are closed under difference, right quotient, reversal, and string homomorphism.
+Regular languages are closed under reversal.
 \end{lem}
@@
 \end{defn}
+
+A proper definition of ${\sf lreplace}$ would give an rewrite system with correctness and termination proofs, which is beyond the scope of this paper
+on escurity-montivated type system exntesions. Instead, we provide an abstract definition of the relation and state necessary propositions.
 
@@
 \begin{proof}
-The theorem follows from closure under difference, right quotient, reversal and string homomorphism.
+Algorithms for the inclusion problem may be adopted to identify any sublanguage $x \subseteq r$ of $r_1$.
+The language $x_{pre}$ can be computed by taking total derivatives until the remaining language
+equals $x$; $x_{post}$ can be computed in a similar way after reversal.
+Then $r'$ is $x_{pre}r_2x_{post}$.
 \end{proof}
@@
 In conclusion, our contribution is a type system, implemented within an extensible type system, for checking the correctness of input sanitation algorithms.
+
+\section{Future Work}
+
+We believe that this type system extension serves as a useful basis for web-oriented static analysis.
+The use sites of arbitrary strings are usually constrained to a few sections of the application. Assuming
+regular expressions are used for input sanitation, annotations at these use sites could be sufficient input.
+
+We also believe that extensible programming languages are a promising approach toward incorporating security
+and privacy into programming languages. The Software Industry moves quickly, and developing new languages for 
+the particular security and privacy concerns of each new domain is likely a losing strategy. Conversely, extensible
+languages might provide developers with a way to incorporate domain-specific security and privacy analyses into their
+daily development activities. In future work, we hope to bolster this hypothesis by extending our approach in this paper
+to more realistic settings and by developing new type extensions addressing security and privacy concerns.
 

### 2014-08-19 sanitation-psp14/sanitation-psp14-short.tex
Added caveat about natural

@@
 In this section, we define a minimal typed lambda calculus with regular string types called $\lambda_{RS}$.
-The syntax of $\lambda_{RS}$ is specified in Figure \ref{fig:lamrs}, its static semantics in Figure \ref{fig:etyping} and its evaluation semantics in Figure \ref{fig:eeval}.
-This will serve as the source language for our translation to a calculus with only standard strings and regular expressions, defined in Sec.
-\ref{translation}.
+The syntax of $\lambda_{RS}$ is specified in Figure \ref{fig:glambdas}, its static semantics in Figure \ref{fig:slambdas} and its evaluation semantics in Figure \ref{fig:dlambdas}.
+This will serve as the source language for our translation to a calculus with only standard strings and regular expressions, defined in Section~\ref{sec:tr}.
 
@@
 
-Type soundness for $\lambdas$ follows from the lemmas and propositions established in the previous subsection.
-
 In this section, we establish type soundness for $\lambdas$.
@@
 by our theorems about regular languages.
+We avoid the most significant issues inherent to big-step semantics by avoiding non-termination.
+The simply typed lambda calculus terminates, and our conservative, compositional extension clearly
+terminates modulo termination of ${\sf subst}$ and ${\sf lreplace}$ (but even if these were non-terminating, this still does not pose a problem for the big-step semantics itself).
+However, a more careful treatment or treatment for a language with non-termination might proceed by either coinduction \cite{coinductionoo, coinduction} or step-indexing \cite{stepindexing}.
 
@@
 correctness of our translation.
+Again, we elide a more careful treatment by treating a special case where our language terminates but
+note the multiple approaches to proving soundness for non-terminating language swith natural semantics \cite{coinductionoo, stepindexing, coinduction}.
 
@@
 \begin{proof}
-  By theorem \ref{thm:tcorrect}, $\iota \sreduces \tstr{s}$ implies that $e \sreduces \strin{s}$.
+  By theorem \ref{thm:trcorrect}, $\iota \sreduces \tstr{s}$ implies that $e \sreduces \strin{s}$.
   By theorem \ref{thm:scorrect}, the above property together with the assumption that $e$ is well-typed implies that $s \in \lang{r}$.

### 2014-08-19 sanitation-psp14/sanitation-psp14-short.tex
Typos

@@
 \item In section \ref{sec:rs}, we describe $\lambdas$. We also prove provide proof outlines for type soundness and a correctness theorem.
-\item In section \ref{sec:p}, we present a simple target language with a minimal regular expression library.
+\item In section \ref{sec:p}, we present a simple target language, $\lambdap$, with a minimal regular expression library.
 \item Section \ref{sec:tr} describes the translation from $\lambdas$ to $\lambdap$ and ensures the correcness result for \ref{sec:rs} is preserved under translation. 
@@
 
-The system $\lambdas$ provides many familiar operations on strings.
+The system $\lambdas$ provides four familiar operations on strings.
 However, unlike most languages, the type system of $\lambdas$ tracks the effect of these operations by relating computations on strings to corresponding computations on regular languages.
+Since these computations terminate, we can use the regular expressions as a basis for static analysis of sanitation protocols.
+
 This section describes our approach, culminating in a Security Theorem.
-The main technical result of this paper will show that this Security Theorem is preserved when $\lambdas$ is translated to a more traditional language.
+The main technical result of this paper shows that this Security Theorem is preserved when $\lambdas$ is translated to a more traditional language.
 Throughout this section, we refer to \textsc{Inference Rules} defined in the figures contained in this section.
@@
 
-Whereas concatenation allows the construction of large strings from smaller strings, \textsc{S-T-Case} allows the decomposition, or elimination, of large strings from smaller strings.
+Whereas concatenation allows the construction of large strings from smaller strings, \textsc{S-T-Case} allows the decomposition, or elimination, of large strings into smaller strings.
 Intuitively, this rule ``peels off" the first character, and then performs a specified operation on the first character and the remaining string.
@@
 For instance, the first character of ``CMU" is always ``C"; however, if the expression associated with this string is CMU$\cup$KIT, then the expression describing the first character is C$\cup$K.
+Therefore, an operation on languages corresponding to case analysis on strings must consider one-character prefixes (a head) and their corresponding suffixes (a tail).
 
@@
 A proper definition of ${\sf lreplace}$ would give an rewrite system with correctness and termination proofs, which is beyond the scope of this paper
-on escurity-montivated type system exntesions. Instead, we provide an abstract definition of the relation and state necessary propositions.
+on security-motivated type system extensions. Instead, we provide an abstract definition of the relation and state necessary propositions.
 
@@
 \begin{prop}[Substitution Correspondence.] \label{thm:substcorrespondence}
-  If $s_1 \in \lang{r_1}$ and $s_2 \in \lang{r_2}$ then $\lsubst{r}{s_1}{s_2} \in \lang{\lreplace{r}{s_1}{s_2}}$.
+  If $s_1 \in \lang{r_1}$ and $s_2 \in \lang{r_2}$ then
+
+  $\lsubst{r}{s_1}{s_2} \in \lang{\lreplace{r}{s_1}{s_2}}$.
 \end{prop}
 \begin{proof}
-The theorem follows from the definitions of subst and lreplace; note that language substitutions over-approximate string substitutions.
+The proposition follows from the definitions of subst and lreplace; note that language substitutions over-approximate string substitutions.
 \end{proof}
@@
 We also rely on additional lemmas which establish the preservation of well-formedness
-of regular epxressions.
+of regular expressions.
 
@@
 \begin{proof}
-  Follows directly from type safety, canonical forms for $\lambdas$, and inversion of the typing relation for $\lambdas$.
+  The theorem follows directly from type safety, canonical forms for $\lambdas$, and inversion of the typing relation for $\lambdas$.
 \end{proof}
@@
 with a $\str$ type and a regular expression type, as well as 
-some operations found in any standard library such as string concatenation and replacement.
+some operations -- such as concatenation and replacement -- found in many the standard libraries of many programming languages.
 The operations of $\lambdap $ correspond precisely to the operations of $\lambdas$ in a way made precise by the translation rules described in the next section.
@@
 Again, we elide a more careful treatment by treating a special case where our language terminates but
-note the multiple approaches to proving soundness for non-terminating language swith natural semantics \cite{coinductionoo, stepindexing, coinduction}.
+note the multiple approaches to proving soundness for non-terminating languages with natural semantics \cite{coinductionoo, stepindexing, coinduction}.
 
@@
 \begin{proof}
-The proof proceeds by induction on the typing relation for $e$ and an appropraite choich of $\iota$; in each case,
+The proof proceeds by induction on the typing relation for $e$ and an appropriate choich of $\iota$; in each case,
 the choice is the syntactic form in $\lambdap$ corresponding to the form under consideration (e.g. choose {\sf preplace} when considering {\sf sreplace}).
@@
 The result of applying sanitize to input is appended in two functions which
-construct a safe query (avoiding command injection) and safe HTML output (avoiding XSS-style attacks).
+construct a safe query (avoiding command injection) and safe HTML output (avoiding Cross-Site Scripting (XSS) attacks).
 The arguments to the result and output construction functions constitute \emph{specifications}.
@@
 In the case of \textsf{results_div}, this specification ensures that user input does 
-not contain any HTML tags, which is a conservative but effective policy for preventing
-Cross-Site Scripting (XSS) attacks.
+not contain any HTML tags, which is a conservative but effective policy for preventing XSS attacks.
 
@@
 
-We implemented a variation on the type system presented in this paper. The only
-significant difference is that we only support replacements where $s_2$ is the empty
+We implemented a variation on the type system presented in this paper with two
+significant differences. first, we only support replacements where $s_2$ is the empty
 string. Therefore, our implementation respects the system presented in section 2
-only modulo the definition of \textsf{lreplace}.
+only modulo the definition of \textsf{lreplace}. Seocnd, we use subtyping instead of
+\textsc{S-*-SafeCoerce}, which decreases inessential verbosity of the language.
 
@@
 The use sites of arbitrary strings are usually constrained to a few sections of the application. Assuming
-regular expressions are used for input sanitation, annotations at these use sites could be sufficient input.
+regular expressions are used for input sanitation, annotations at these use sites could be sufficient input to a sound and complete static analysis.
 

### 2014-08-19 sanitation-psp14/sanitation-psp14-short.tex
Updated description.

@@
 
- \textsc{S-T-SafeCoerce}: \todo{fix this description} allows only safe coercions between string types by
-    expoiting the decidability of language inclusion.
-
-\textsc{S-T-Check}: \todo{fix this description}The check rules allow casts between strings that cannot be guaranteed at compile time, but might pass a runtime check.
-
-\subsubsection{Replacement}
+The $\lambdas$ language supports two forms of coercion. Safe coercion only allows coercsion between 
+strings types which are guaranteed to be safe.
+This form of coercion is useful because the replacement operation (introduced below in Section~\ref{sec:replace}) is often more conservative than absolutely necessary.
+Conversely, unsafe coercion -- which we refer to as a checked coercion -- 
+allows for potential unsafe type casts. Our semantics for check ensures that only safe values are used. Whenever a value is determined to be unsafe at runtime, a default value $e_2$ is used instead.
+
+Summarily, 
+ \textsc{S-T-SafeCoerce} allows only safe coercions between string types by
+    expoiting the decidability of language inclusion, while
+\textsc{S-T-Check} allows casts between strings that cannot be guaranteed at compile time, but inserts a runtime check a runtime check.
+
+\subsubsection{Replacement}\label{sec:replace}
 

### 2014-08-19 sanitation-psp14/sanitation-psp14-short.tex
\cup -> +

@@
 However, a regular expression describing the first character of any matching string might recognize multiple concrete values.
-For instance, the first character of ``CMU" is always ``C"; however, if the expression associated with this string is CMU$\cup$KIT, then the expression describing the first character is C$\cup$K.
+For instance, the first character of ``CMU" is always ``C"; however, if the expression associated with this string is CMU$+$KIT, then the expression describing the first character is C$+$K.
 Therefore, an operation on languages corresponding to case analysis on strings must consider one-character prefixes (a head) and their corresponding suffixes (a tail).
@@
 \item $\lhead{a r'} = a$ where $a \in \Sigma$
-\item $\lhead{r_1 \cup  r_2}  = \lhead{r_1} \cup \lhead{r_2}$
+\item $\lhead{r_1 +  r_2}  = \lhead{r_1} + \lhead{r_2}$
 \item $\lhead{r*} = \lhead{r}$
-\item $\lhead{.} = a_1 \cup a_2 \cup ... \cup a_n$ for all $a_i \in \Sigma$ where $|\Sigma| = n$
+\item $\lhead{.} = a_1 + a_2 + ... + a_n$ for all $a_i \in \Sigma$ where $|\Sigma| = n$
 \item $\lhead{\epsilon} = \epsilon$.
@@
 The function \ltail{r} is defined in terms of $\lhead{r}$.
-Note that $\lhead{r} = a_1 \cup a_2 \cup ... \cup a_i$.
-We define $\ltail{r} = \delta_{a_1}(r) \cup \delta_{a_2}(r) \cup ... \cup \delta_{a_i}(r)$
+Note that $\lhead{r} = a_1 + a_2 + ... + a_i$.
+We define $\ltail{r} = \delta_{a_1}(r) + \delta_{a_2}(r) + ... + \delta_{a_i}(r)$
 where $\delta_s(r)$ is the Brzozowski derivative of $r$ with respect to $s$ \cite{bowzer}.

### 2014-08-19 sanitation-psp14/sanitation-psp14-short.tex
proof reading (~jw)

@@
 To be practical, adopting a specialized type system like this should not require the adoption of a new programming language.
-Instead, we favor extensible type systems: new type system fragments like this should be implemented as libraries atop a mechanism that guarantees that they can be safely composed.
-We support this by 1) specifying a translation from $\lambda_{RS}$ to a language containing only strings and regular expressions, then, taking Python as such a language, 2) implement the type system together with the translation as a library using \texttt{atlang}, an extensible static type system for Python (being developed by the authors).% represents the most sophisticated application of this library to date.
+Instead, we advocate for extensible type systems: new type system fragments like this should be implemented as libraries atop a mechanism that guarantees that they can be safely composed.
+We support this by 1) specifying a translation from $\lambda_{RS}$ to a language containing only strings and regular expressions, then, taking Python as such a language, 2) implementing the type system together with the translation as a library using \texttt{atlang}, an extensible static type system for Python (being developed by the authors).% represents the most sophisticated application of this library to date.
 

### 2014-08-19 sanitation-psp14/sanitation-psp14-short.tex
small changes to beginning of section 2

@@
 The syntax of $\lambda_{RS}$ is specified in Figure \ref{fig:glambdas}, its static semantics in Figure \ref{fig:slambdas} and its evaluation semantics in Figure \ref{fig:dlambdas}.
-This will serve as the source language for our translation to a calculus with only standard strings and regular expressions, defined in Section~\ref{sec:tr}.
-
+%This will serve as the source language for our translation to a calculus with only standard strings and regular expressions, defined in Section~\ref{sec:tr}.
+
+An outline for this section follows:
+\begin{itemize}
+\item Section \ref{sec:rs} describes these rules. We also give proof outlines of  type safety and correctness theorems.
+\item Section \ref{sec:p} describes a simple target language, $\lambdap$, with a minimal regular expression library. In Section \ref{atlang}, we will take Python to be such a language.
+\item Section \ref{sec:tr} describes the translation from $\lambdas$ to $\lambdap$ and ensures the correcness result for \ref{sec:rs} is preserved under translation. 
+\end{itemize}
+
+\subsection{The Language of Regular Strings}\label{sec:rs}
 There are two type constructors in $\lambdas$, $\rightarrow$ and $\textsf{stringin}$.
@@
 $s \in \Sigma^\star$, defined in the usual way).
-The rule \textsc{S-T-Stringin-I} ensures that $s \in \lang{r}$.
-The remaining operations on terms of type $\stringin{r}$ preserve this property.
-%The remaining operators operate on regular strings.
-
-An outline for this section follows:
-\begin{itemize}
-\item In section \ref{sec:rs}, we describe $\lambdas$. We also prove provide proof outlines for type soundness and a correctness theorem.
-\item In section \ref{sec:p}, we present a simple target language, $\lambdap$, with a minimal regular expression library.
-\item Section \ref{sec:tr} describes the translation from $\lambdas$ to $\lambdap$ and ensures the correcness result for \ref{sec:rs} is preserved under translation. 
-\end{itemize}
-
-\subsection{The Language of Constrained Strings}\label{sec:rs}
-
-The system $\lambdas$ provides four familiar operations on strings.
-However, unlike most languages, the type system of $\lambdas$ tracks the effect of these operations by relating computations on strings to corresponding computations on regular languages.
-Since these computations terminate, we can use the regular expressions as a basis for static analysis of sanitation protocols.
-
-This section describes our approach, culminating in a Security Theorem.
+The rule \textsc{S-T-Stringin-I} statically guarantees that $s \in \lang{r}$.\footnote{For convenience, a single sheet containing Figures \ref{fig:glambdas}-\ref{fig:tr} is available at http://nfulton.org/strings/printout.pdf.}.
+ %The remaining operations on terms of type $\stringin{r}$ preserve this property. %The remaining operators operate on regular strings.
+
+$\lambdas$ provides several familiar operations on strings.
+The type system  handles these operations by relating operations over strings to corresponding operations over the  regular languages they belong to.
+Since these operations over regular languages are known to be closed and decidable, we can use the corresponding operations over regular strings as a basis for static analysis of sanitation protocols.
+
+%This section describes our approach, culminating in a Security Theorem.
 The main technical result of this paper shows that this Security Theorem is preserved when $\lambdas$ is translated to a more traditional language.
 Throughout this section, we refer to \textsc{Inference Rules} defined in the figures contained in this section.
-\todo{remove tranditional, most and improve description}
-
-In this section, we define $\lambdas$, describe each operation, and finally prove both type safety and a correctness result for the language.
-The grammar for $\lambdas$ is defined in Figure \ref{fig:glambdas}. Typing rules \textsc{S-T-} are defined in Figure \ref{fig:slambdas} and a big-step semantics
-is defined in Figure \ref{fig:dlambdas}
-\footnote{For the reader's convienance, a single sheet containing figures \ref{fig:glambdas}-\ref{fig:tr} is available at http://nfulton.org/strings/printout.pdf.}.
+%\todo{remove tranditional, most and improve description}
+
+%In this section, we define $\lambdas$, describe each operation, and finally prove both type safety and a correctness result for the language.
+%The grammar for $\lambdas$ is defined in Figure \ref{fig:glambdas}. Typing rules \textsc{S-T-} are defined in Figure \ref{fig:slambdas} and a big-step semantics
+%is defined in Figure \ref{fig:dlambdas}
 

### 2014-08-19 sanitation-psp14/sanitation-psp14-short.tex
small changes to sec2

@@
 \section{Regular String Types, Minimally}\label{calculus}
-
-In this section, we define a minimal typed lambda calculus with regular string types called $\lambda_{RS}$.
-The syntax of $\lambda_{RS}$ is specified in Figure \ref{fig:glambdas}, its static semantics in Figure \ref{fig:slambdas} and its evaluation semantics in Figure \ref{fig:dlambdas}.
+\noindent
 %This will serve as the source language for our translation to a calculus with only standard strings and regular expressions, defined in Section~\ref{sec:tr}. 
-
-An outline for this section follows:
+This section is organized as follows:
 \begin{itemize}
-\item Section \ref{sec:rs} describes these rules. We also give proof outlines of  type safety and correctness theorems.
+\item Section \ref{sec:rs} describes $\lambda_{RS}$. We also give proof outlines of  type safety and correctness theorems.
 \item Section \ref{sec:p} describes a simple target language, $\lambdap$, with a minimal regular expression library. In Section \ref{atlang}, we will take Python to be such a language.
@@
 \subsection{The Language of Regular Strings}\label{sec:rs}
+In this section, we define a minimal typed lambda calculus with regular string types called $\lambda_{RS}$.
+The syntax of $\lambda_{RS}$ is specified in Figure \ref{fig:glambdas}, its static semantics in Figure \ref{fig:slambdas} and its evaluation semantics in Figure \ref{fig:dlambdas}.\footnote{For convenience, a single sheet containing Figures \ref{fig:glambdas}-\ref{fig:tr} is available at http://nfulton.org/strings/printout.pdf.}
+
 There are two type constructors in $\lambdas$, $\rightarrow$ and $\textsf{stringin}$.
@@
 $s \in \Sigma^\star$, defined in the usual way).
-The rule \textsc{S-T-Stringin-I} statically guarantees that $s \in \lang{r}$.\footnote{For convenience, a single sheet containing Figures \ref{fig:glambdas}-\ref{fig:tr} is available at http://nfulton.org/strings/printout.pdf.}.
+The rule \textsc{S-T-Stringin-I} statically guarantees that $s \in \lang{r}$.
  %The remaining operations on terms of type $\stringin{r}$ preserve this property. %The remaining operators operate on regular strings.

### 2014-08-20 sanitation-psp14/sanitation-psp14-short.tex
small changes to sec2
Minor changes
added values.
values.
Updated intro

@@
 
-\newcommand{\strcase}[3]{ {\sf strcase}(#1; #2; #3)}
+\newcommand{\strcase}[3]{ {\sf rstrcase}(#1; #2; #3)}
 \newcommand{\pstrcase}[3]{ {\sf pstrcase}(#1; #2; #3)}
@@
 For example, web browsers are controlled using HTML and Javascript sent from a server as a string, and database engines execute SQL queries also sent as strings.
-When these commands include data derived from user input, care must be taken to ensure that the user cannot  subverts the intended command by carefully crafting the data they send.
+When these commands include data derived from user input, care must be taken to ensure that the user cannot  subvert the intended command by carefully crafting the data they send.
 For example, a  SQL query constructed using string concatenation exposes a SQL injection vulnerability:
@@
 For example, in this case, the developer (or, more often, a framework) might define a function \verb|sanitize| that escapes double quotes and existing backslashes   with a backslash, which SQL treats safely.
-Guaranteeing that user input has already been sanitized like this before it is used to construct a command is challenging.
-Note that this function is not idempotent, so it should only be called once.
+Alternatively, it might HTML-encode special characters, which would head off both SQL injection attacks and cross-site scripting attacks.
+Guaranteeing that user input has already been sanitized  before it is used to construct a command is challenging, because  sanitation is often performed in a different part of the code than the place where commands are ultimately constructed.% Note, for example, that \verb|sanitize| is not idempotent, so it should only be called once.
 
@@
 In this paper, we present a static type system that tracks the regular language a string belongs to.
-For example, the output of \verb|sanitize| will be a string in the regular language described by the regular expression above (we describe such a regular language, following convention, as $\mathcal{L}\{r\}$).
-By leveraging closure and decidability properties of regular languages, the language of a string is tracked through uses of a number of operations, including \emph{replacement} of substrings matching a given regular expression.
-This makes it simple to implement sanitation procedures like the one just described in an essentially conventional manner.
+For example, the output of \verb|sanitize| will be a string in the regular language described by the regular expression above (we identify regular languages by the notation $\mathcal{L}\{r\}$).
+By leveraging closure and decidability properties of regular languages, the type system tracks the language of a string through uses of a number of operations, including \emph{replacement} of substrings matching a given pattern.
+This makes it possible to implement sanitation functions like the one just described in an essentially conventional manner.
 The result is a system where the fact that a string has been {correctly} sanitized is manifest in its type.
@@
 These guarantees require run-time checks only when going from less precise to more precise types (e.g.
-at the edges of the system, where user input has not yet been validated).
+at the edges of the system).
 %implementing and statically checking input sanitation techniques.
@@
 Coercions that are known to be  safe due to a sublanguage relationship are performed implicitly, also without run-time overhead.
-This results in a \emph{usably secure} system: working with regular strings differs little from working with standard strings.
+This results in a \emph{usably secure} system: working with regular strings differs little from working with standard strings in a language that web developers have already widely adopted.
 
@@
 \begin{itemize}
-\item Section \ref{sec:rs} describes $\lambda_{RS}$. We also give proof outlines of  type safety and correctness theorems.
+\item Section \ref{sec:rs} describes $\lambda_{RS}$. We also give proof outlines of  type safety and correctness theorems and relevant propositions about regular languages.
 \item Section \ref{sec:p} describes a simple target language, $\lambdap$, with a minimal regular expression library. In Section \ref{atlang}, we will take Python to be such a language.
@@
 
-$\lambdas$ provides several familiar operations on strings.
-The type system  handles these operations by relating operations over strings to corresponding operations over the  regular languages they belong to.
-Since these operations over regular languages are known to be closed and decidable, we can use the corresponding operations over regular strings as a basis for static analysis of sanitation protocols.
+$\lambdas$ provides several familiar operations on strings. The type system  relates these operations over strings to corresponding operations over the  regular languages they belong to. 
+Since these operations over regular languages are known to be closed and decidable, we can use these operations as a basis for static analysis of sanitation protocols (we will see a concrete example in Section \ref{atlang}).
 
@@
       $\rsconcat{e}{e}$ \alt
+      $\strcase{e}{e}{x,y.e}$ \alt
       $\sisubst{r}{e}{e}$ \alt
@@
 
-<P> ::= $x$  \hfill target terms \alt
+<$\iota$> ::= $x$  \hfill target terms \alt
    $\lambda x . \iota$ \alt
@@
   $\tconcat{\iota}{\iota}$ \alt
+  $\pstrcase{\iota}{\iota}{x.y.\iota}$ \alt
   $\preplace{\iota}{\iota}{\iota}$ \alt
@@
 
-$\fbox{\inferrule{}{\sctx \vdash \trden{e} = \iota}}$
+$\fbox{\inferrule{}{ \trden{e} = \iota}}$
 \begin{mathpar}
@@
 { }
-{\sctx \vdash \trden{x} = x}
+{ \trden{x} = x}
 
 \inferrule[Tr-Abs]
-{ \sctx \vdash \trden{e} = \iota }
-{\sctx \vdash \trden{ \lambda x.e } = \lambda x . \iota}
+{  \trden{e} = \iota }
+{ \trden{ \lambda x.e } = \lambda x . \iota}
 
 \inferrule[Tr-App]
-{\sctx \vdash \trden{e_1} = \iota_1 \\ \sctx \vdash \trden{e_2} = \iota_2}
-{\sctx \vdash \trden{e_1(e_2)} = \iota_1(\iota_2)}
+{ \trden{e_1} = \iota_1 \\  \trden{e_2} = \iota_2}
+{ \trden{e_1(e_2)} = \iota_1(\iota_2)}
 
 \inferrule[Tr-Case]
-{\sctx \vdash \trden{e_1} = \iota_1 \\
-  \sctx \vdash \trden{e_2} = \iota_2 \\
-  \sctx \vdash \trden{e_3} = \iota_3
-}
-{\sctx \vdash \trden{ \strcase{e_1}{e_2}{e_3} } = \pstrcase{\iota_1}{\iota_2}{\iota_3} }
+{ \trden{e_1} = \iota_1 \\
+   \trden{e_2} = \iota_2 \\
+   \trden{e_3} = \iota_3
+}
+{ \trden{ \strcase{e_1}{e_2}{e_3} } = \pstrcase{\iota_1}{\iota_2}{\iota_3} }
 
@@
 { \ }
-{ \sctx \vdash  \trden{\strin{s}} = \tstr{s}}
+{   \trden{\strin{s}} = \tstr{s}}
 
 \inferrule[Tr-Concat]
-{ \sctx \vdash \trden{e_1} = \iota_1 \\ \sctx \vdash \trden{e_2} = \iota_2}
-{ \sctx \vdash \trden{\rsconcat{e_1}{e_2}} = \tconcat{\iota_1}{\iota_2} }
+{  \trden{e_1} = \iota_1 \\  \trden{e_2} = \iota_2}
+{  \trden{\rsconcat{e_1}{e_2}} = \tconcat{\iota_1}{\iota_2} }
 
 \inferrule[Tr-Subst]
-{ \sctx \vdash  \trden{e_1} = \iota_1 \\ \sctx \vdash  \trden{e_2} = \iota_2 }
-{ \sctx \vdash  \trden{ \sisubst{r}{e_1}{e_2} } = \tsubst{\rx{r}}{\iota_1}{\iota_2} }
+{   \trden{e_1} = \iota_1 \\   \trden{e_2} = \iota_2 }
+{   \trden{ \sisubst{r}{e_1}{e_2} } = \tsubst{\rx{r}}{\iota_1}{\iota_2} }
 
 \inferrule[Tr-SafeCoerce]
-{ \sctx \vdash \trden{e} = \iota }
-{ \sctx \vdash  \trden{ \coerce{r'}{e} } = \iota }
+{  \trden{e} = \iota }
+{   \trden{ \coerce{r'}{e} } = \iota }
 
 \inferrule[Tr-Check]
-{ \sctx \vdash \trden{e} = \iota_1 \\ \sctx \vdash \trden{e_1} = \iota_1?? \\ \sctx \vdash \trden{e_2} = \iota_2} 
-{ \sctx \vdash \trden{ \rcheck{r}{e}{x.e_1}{e_2} } = \tcheck{\rx{r}}{\iota}{x.\iota_1}{\iota_2} }
+{  \trden{e} = \iota_1 \\  \trden{e_1} = \iota_1 \\  \trden{e_2} = \iota_2} 
+{  \trden{ \rcheck{r}{e}{x.e_1}{e_2} } = \tcheck{\rx{r}}{\iota}{x.\iota_1}{\iota_2} }
 

### 2014-08-20 sanitation-psp14/sanitation-psp14-short.tex
fixed fig2
fixes to fig4
s-e-case-concat cleanup
simultaneous substitution; use values instead of bare strings.
Cleaned up end of mathpar.
removed binder from pcheck.
More figure cleanup.
Fixed fig 4
newline.
tr-case
fig8 is no longer type directed.
Added definition of \delta_a(r)

@@
 \renewcommand{\tstr}[1]{{{\sf str}[#1]}}
-\newcommand{\preplace}[3]{{\sf preplace}(#1;#2;#3)}
+\newcommand{\preplace}[3]{{\sf replace}(#1;#2;#3)}
 \newcommand{\tconcat}[2]{{\sf concat}(#1;#2)} \newcommand{\concat}[2]{{\sf concat}(#1;#2)} % lots of legacy naming around these parts...
@@
 \newcommand{\strcase}[3]{ {\sf rstrcase}(#1; #2; #3)}
-\newcommand{\pstrcase}[3]{ {\sf pstrcase}(#1; #2; #3)}
+\newcommand{\pstrcase}[3]{ {\sf strcase}(#1; #2; #3)}
 
@@
       $\rcheck{r}{e}{x.e}{e}$
+
+<$v$> ::= $\lambda x . e$ \hfill source values \alt
+          $\strin{s}$ \hfill $s \in \Sigma^{*}$
+
 \caption{Syntax of $\lambda_{RS}$.}
@@
 {
-  \sctx \vdash \strcase{e_1}{e_2}{x.y.e_3} : \sigma
+  \sctx \vdash \strcase{e_1}{e_2}{x,y.e_3} : \sigma
 }
@@
 {\sctx \vdash \rcheck{r}{e_0}{x.e_1}{e_2} : \sigma}
-
 \end{mathpar}
@@
   \small
-$\fbox{\inferrule{}{e \sreduces e}}$
+$\fbox{\inferrule{}{e \sreduces v}}$
 \begin{mathpar}
@@
 \inferrule[S-E-App]
-{ e_1 \sreduces \lambda x . e_3 \\ e_2 \sreduces e_2' \\ [ e_2' / x ] e_3 \sreduces v }
+{ e_1 \sreduces \lambda x . e_3 \\ e_2 \sreduces v_2 \\ [ v_2 / x ] e_3 \sreduces v }
 { e_1(e_2) \sreduces v }
@@
 {
-  \strcase{e_1}{e_2}{e_3} \sreduces v_2
+  \strcase{e_1}{e_2}{x,y.e_3} \sreduces v_2
 }
@@
 {
-  e_1 \sreduces \strin{p s} \\
-  e_3 \sreduces x.y.e_4 \\
-  [p/x][s/y]e_4 \sreduces v
+  e_1 \sreduces \strin{a s} \\
+  [\rstr{a},\rstr{s} / x,y]e_3 \sreduces v_3
 }
 {
-  \strcase{e_1}{e_2}{e_3} \sreduces v
+  \strcase{e_1}{e_2}{x,y.e_3} \sreduces v_3
 }
@@
 \inferrule[S-E-Replace]
-{e_1 \sreduces \strin{e_1} \\ e_2 \sreduces \strin{s_2} \\ \lsubst{r}{s_1}{s_2} = s} 
+{e_1 \sreduces \strin{s_1} \\ e_2 \sreduces \strin{s_2} \\ \lsubst{r}{s_1}{s_2} = s} 
 {\sisubst{r}{e_1}{e_2} \sreduces \sistr{s}}
@@
 {
-  e \sreduces \strin{s} \\ s \not \in \lang{r}
+  e \sreduces \strin{s} \\ s \not \in \lang{r} \\
+  e_2 \sreduces v
 }
 {
-  \rcheck{r}{e}{x.e_1}{e_2} \sreduces e_2
-}
-
+  \rcheck{r}{e}{x.e_1}{e_2} \sreduces v
+}
 \end{mathpar}
@@
 \item $\lhead{a r'} = a$ where $a \in \Sigma$
+\item $\lhead{\epsilon} = \epsilon$
 \item $\lhead{r_1 +  r_2}  = \lhead{r_1} + \lhead{r_2}$
 \item $\lhead{r*} = \lhead{r}$
-\item $\lhead{.} = a_1 + a_2 + ... + a_n$ for all $a_i \in \Sigma$ where $|\Sigma| = n$
-\item $\lhead{\epsilon} = \epsilon$.
+\item $\lhead{.} = a_1 + a_2 + ... + a_n$ for all $a_i \in \Sigma$ where $|\Sigma| = n$.
 \end{itemize}
@@
 Given this definition of $\lhead{r}$, regular expression derivatives provide a useful tool for defining $\ltail{r}$.
+\begin{defn}[Brzozowski's Derivative]\label{def:derivative}
+  The \emph{derivative of $r$ with respect to $s$} is denoted by $\delta_s(r)$
+  and is $\delta_s(r) = \{t | st \in \lang{r}\}$.
+\end{defn}
+
+Definition~\ref{def:derivative} is equivalent to the definition given in \cite{bowzer},
+although we refer the unfamiliar reader to \cite{owens}.
+Definition~\ref{def:derivative} is equivalent to Definition 3.1 in both papers.
+We now define $\ltail{r}$ using derivatives with respect to $\lhead{r}$.
 
@@
 Note that $\lhead{r} = a_1 + a_2 + ... + a_i$.
-We define $\ltail{r} = \delta_{a_1}(r) + \delta_{a_2}(r) + ... + \delta_{a_i}(r)$
-where $\delta_s(r)$ is the Brzozowski derivative of $r$ with respect to $s$ \cite{bowzer}.
+We define $\ltail{r} = \delta_{a_1}(r) + \delta_{a_2}(r) + ... + \delta_{a_i}(r)$.
 \end{defn}
@@
              $\lambda x . \iota$ \alt
-   $\iota \iota$ \alt
+                    $\iota(\iota)$ \alt
 $\tstr{s}$ \alt
@@
   $\tconcat{\iota}{\iota}$ \alt
-  $\pstrcase{\iota}{\iota}{x.y.\iota}$ \alt
+  $\pstrcase{\iota}{\iota}{x,y.\iota}$ \alt
   $\preplace{\iota}{\iota}{\iota}$ \alt
-  $\tcheck{\rx{r}}{\iota}{\iota}{\iota}$ 
+  $\tcheck{\iota}{\iota}{\iota}{\iota}$
+
+<$\dot{v}$> ::= $\lambda x . \iota$ \hfill target values \alt
+                $\tstr{s}$ \alt
+                $\rx{r}$
+
 \end{grammar}
@@
 {\tctx \vdash \iota_1 : \str \\ \tctx \vdash \iota_2 : \tau \\ \tctx, x:\str, y:\str \vdash \iota_3:\tau}
-{\tctx \vdash \pstrcase{\iota_1}{\iota_2}{\iota_3} : \tau}
+{\tctx \vdash \pstrcase{\iota_1}{\iota_2}{x,y.\iota_3} : \tau}
 
@@
 \inferrule[P-T-Check]
-{\tctx \vdash \iota_r : \regex \\ \tctx \vdash \iota_1 : \str \\ \tctx, x:\str \vdash \iota_2 : \sigma \\ \tctx \vdash \iota_3 : \sigma}
-{\tctx \vdash \tcheck{\iota_r}{\iota_1}{x . \iota_2}{\iota_3} : \sigma}
+{\tctx \vdash \iota_r : \regex \\ \tctx \vdash \iota_1 : \str \\ \tctx \vdash \iota_2 : \sigma \\ \tctx \vdash \iota_3 : \sigma}
+{\tctx \vdash \tcheck{\iota_r}{\iota_1}{\iota_2}{\iota_3} : \sigma}
 \end{mathpar}
@@
 \small
-$\fbox{\inferrule{}{\iota \treduces \iota}}$
+$\fbox{\inferrule{}{\iota \treduces \dot{v}}}$
 
@@
 \inferrule[P-E-App]
-{ \iota_1 \sreduces \lambda x . \iota_3 \\  \iota_2 \sreduces \iota_2' \\ [\iota_3 / x] \iota_2' \sreduces v}
-{ \iota_1(\iota_2) \sreduces v}
+{ \iota_1 \sreduces \lambda x . \iota_3 \\  \iota_2 \sreduces \dot{v}_2 \\ [\dot{v}_2 / x] \iota_3 \sreduces \dot{v}_3}
+{ \iota_1(\iota_2) \sreduces \dot{v}_3}
 
@@
 {
-  \iota_1 \treduces \tstr{\ } \\
-  \iota2 \treduces \iota_2
+  \iota_1 \treduces \tstr{ \epsilon } \\
+  \iota_2 \treduces \dot{v_2} \\
 }
 {
-  \pstrcase{\iota_1}{\iota_2}{\iota_3} \treduces v_2
+  \pstrcase{\iota_1}{\iota_2}{x,y.\iota_3} \treduces \dot{v_2}
 }
@@
 {
-  \iota_1 \treduces \tstr{ps} \\
-  x.y.\iota_3 \treduces \iota_4 \\
-  [p/x][s/y]\iota_4 \treduces v
+  \iota_1 \treduces \tstr{as} \\
+  [\tstr{a},\tstr{s}/x,y] \iota_3 \treduces \dot{v} 
 }
 {
-  \pstrcase{\iota_1}{\iota_2}{\iota_3} \treduces v
+  \pstrcase{\iota_1}{\iota_2}{x,y.\iota_3} \treduces \dot{v}
 }
@@
 \inferrule[P-E-Check-OK]
-{\iota \treduces \tstr{s} \\ s \in \lang{r} \\ [\tstr{s} / x] \iota_1 \sreduces \iota_3}
-{\tcheck{\rx{r}}{\iota}{x.\iota_1}{\iota_2} \treduces \iota_3}
+{
+  \iota_r \treduces \rx{r} \\
+  \iota \treduces \tstr{s} \\ 
+s \in \lang{r} \\ 
+\iota_1 \treduces \dot{v_1} 
+}
+{
+  \tcheck{\iota_r}{\iota}{\iota_1}{\iota_2} \treduces \dot{v_1}
+}
 
 \inferrule[P-E-Check-NotOK]
-{\iota \treduces \tstr{s} \\ s \not \in \lang{r}}
-{\tcheck{\rx{r}}{\iota}{x.\iota_1}{\iota_2} \treduces \iota_2}
+{\iota_r \treduces \rx{r} \\ \iota \treduces \tstr{s} \\ s \not \in \lang{r} \\ \iota_2 \treduces \dot{v_2}}
+{\tcheck{\iota_r}{\iota}{\iota_1}{\iota_2} \treduces \dot{v_2}}
 \end{mathpar}
@@
 }
-{ \trden{ \strcase{e_1}{e_2}{e_3} } = \pstrcase{\iota_1}{\iota_2}{\iota_3} }
+{ \trden{ \strcase{e_1}{e_2}{x,y.e_3} } = \pstrcase{\iota_1}{\iota_2}{x,y.\iota_3} }
 
@@
 \inferrule[Tr-Check]
-{  \trden{e} = \iota_1 \\  \trden{e_1} = \iota_1 \\  \trden{e_2} = \iota_2} 
-{  \trden{ \rcheck{r}{e}{x.e_1}{e_2} } = \tcheck{\rx{r}}{\iota}{x.\iota_1}{\iota_2} }
-
+{  
+  \trden{e} = \iota \\  
+   \trden{e_1} = \iota_1 \\  
+   \trden{e_2} = \iota_2
+} 
+{ 
+  \trden{ \rcheck{r}{e}{x.e_1}{e_2} } = \tcheck{\rx{r}}{\iota}{(\lambda x . \iota_1)(\iota)}{\iota_2} 
+} 
 \end{mathpar}
-\caption{Translation from source terms ($e$) to target terms ($\iota$).
-The translation is type-directed.}
+\caption{Translation from source terms ($e$) to target terms ($\iota$).}
 \label{fig:tr}
@@
 The proof proceeds by induction on the typing relation for $e$ and an appropriate choich of $\iota$; in each case,
-the choice is the syntactic form in $\lambdap$ corresponding to the form under consideration (e.g. choose {\sf preplace} when considering {\sf sreplace}).
+the choice is the syntactic form in $\lambdap$ corresponding to the form under consideration (e.g. choose {\sf replace} when considering {\sf sreplace}).
 After the correct choice, the proof proceeds by our type safety theorems and an appeal to the induction hypothesis.

### 2014-08-20 sanitation-psp14/sanitation-psp14-short.tex
Added contexts.

@@
 
-$\fbox{\inferrule{}{ \trden{e} = \iota}}$
+  $\fbox{\inferrule{}{ \trden{e} = \iota \\ \trden{\sigma} = \tau}}$
 \begin{mathpar}
+\inferrule[Tr-T-String]
+{ }
+{ \trden{\stringin{r}} = \str }
+
+\inferrule[Tr-T-Arrow]
+{ \trden{\sigma_1} = \tau_1 \\ \trden{\sigma_2} = \tau_2 }
+{ \trden{\sigma_1 \rightarrow \sigma_2} = \tau_1 \rightarrow \tau_2 }
+
+\inferrule[Tr-T-Set]
+{ \trden{\sigma_1} = \tau_1 \\  ... \\ \trden{\sigma_n} = \tau_n}
+{ \trden{ \{\sigma_1, \sigma_2 \} } = \{ \tau_1, \tau_2 \} }
+
 \inferrule[Tr-Var]
@@
 \begin{thm}[Translation Correctness]\label{thm:trcorrect}
-  If $\tctx \vdash e : \stringin{r}$ then there exists an $\iota$ such that $\trden{e} = \iota$, $\iota \sreduces \tstr{s}$,
-  and $e \treduces \rstr{s}$.
+  If $\tctx \vdash e : \sigma$ then 
+  there exists an $\iota$ such that $\trden{e} = \iota$
+  and $\trden{\tctx} \vdash \iota : \trden{\sigma}$.
+  Furthermore, $e \sreduces v$ and
+  $\iota \treduces \dot{v}$ such that
+  $\trden{v} = \dot{v}$.
 \end{thm}

### 2014-08-21 sanitation-psp14/sanitation-psp14-short.tex
ctx.
merged

@@
 $\lambdas$ provides several familiar operations on strings. The type system  relates these operations over strings to corresponding operations over the  regular languages they belong to. 
-Since these operations over regular languages are known to be closed and decidable, we can use these operations as a basis for static analysis of sanitation protocols (we will see a concrete example in Section \ref{atlang}).
+Since these operations over regular languages are known to be closed and decidable, we can use these operations as a basis for static analysis of sanitation protocols (we will see a complete example in Section \ref{atlang}).
 
@@
 
-\subsection{A Target Language with a Regular Expression Library}\label{sec:p}
-
-The system $\lambdap$ is a straight-forward extension of a simply typed lambda calculus
-with a $\str$ type and a regular expression type, as well as 
-some operations -- such as concatenation and replacement -- found in many the standard libraries of many programming languages.
-The operations of $\lambdap $ correspond precisely to the operations of $\lambdas$ in a way made precise by the translation rules described in the next section.
-Unlike $\lambdas$, $\lambdap$ does not statically track the effects of string operations.
-The language $\lambdap$ is so-called because it is reminscent of popular web programming languages,
-such as \textbf{P}ython or \textbf{P}HP.
-
-The grammar of $\lambdap$ is defined in Figure~\ref{fig:lcsSyntax}.
-The typing rules \textsc{P-T-} are defined in Figure~\ref{fig:slambdap}
-and a big-step semantics is defined by the rules \textsc{P-E-} in Figure~\ref{fig:dlambdap}.
-
-\renewcommand{\grammarlabel}[2]{#1\hfill#2}
-
-\begin{figure}[h]
+\subsection{Target Language}\label{sec:p}
+\begin{figure}[t]
 \begin{grammar}
 
-<$\theta$> ::= $\theta \rightarrow \theta$ \hfill target types \alt
+<$\tau$> ::= $\tau \rightarrow \tau$ \hfill target types \alt
 $\str$ \alt $\regex$
@@
 
-\begin{figure}[h]\label{fig:lambdap}
+The system $\lambdap$ is a straight-forward extension of a simply typed lambda calculus
+with a $\str$ type and a regular expression type, as well as 
+some operations -- such as concatenation and replacement -- found in many the standard libraries of many programming languages.
+The operations of $\lambdap $ correspond precisely to the operations of $\lambdas$ in a way made precise by the translation rules described in the next section.
+Unlike $\lambdas$, $\lambdap$ does not statically track the effects of string operations.
+The language $\lambdap$ is so-called because it is reminscent of popular web programming languages,
+such as \textbf{P}ython or \textbf{P}HP.
+
+The grammar of $\lambdap$ is defined in Figure~\ref{fig:lcsSyntax}.
+The typing rules \textsc{P-T-} are defined in Figure~\ref{fig:slambdap}
+and a big-step semantics is defined by the rules \textsc{P-E-} in Figure~\ref{fig:dlambdap}.
+
+\renewcommand{\grammarlabel}[2]{#1\hfill#2}
+
+\begin{figure}[t]\label{fig:lambdap}
 \small
@@
 \end{mathpar}
-\caption{Big step semantics for of $\lambdap$}.
+\caption{Big step semantics for $\lambdap$}.
 \label{fig:dlambdap}
@@
 \begin{figure}[h]
-
-  $\fbox{\inferrule{}{ \trden{e} = \iota \\ \trden{\sigma} = \tau}}$
+$\fbox{\inferrule{}{\trden{\sigma}=\tau}}$
 \begin{mathpar}
@@
 { \trden{\sigma_1 \rightarrow \sigma_2} = \tau_1 \rightarrow \tau_2 }
-
-\inferrule[Tr-T-Set]
-{ \trden{\sigma_1} = \tau_1 \\  ... \\ \trden{\sigma_n} = \tau_n}
-{ \trden{ \{\sigma_1, \sigma_2 \} } = \{ \tau_1, \tau_2 \} }
-
+\end{mathpar}
+$\fbox{\inferrule{}{\trden{\Psi}=\Theta}}$
+\begin{mathpar}
+\inferrule[Tr-T-Context-Emp]
+{ }{\trden{\emptyset} = \emptyset}
+
+\inferrule[Tr-T-Context-Ext]
+{\trden{\Psi}=\Theta\\ \trden{\sigma} = \tau}
+{\trden{\Psi, x : \sigma} = \Theta, x : \tau}
+\end{mathpar}
+  $\fbox{\inferrule{}{ \trden{e} = \iota }}$
+\begin{mathpar}
 \inferrule[Tr-Var]
@@
 In this section, we take Python to be such a target language.
-Python does not have a static type system, however, so to implement these semantics, we will leverage \verb|atlang|, an extensible type system for Python (being developed by the authors).
-By using \verb|atlang|, which builds on Python's quotations and reflection facilities, we can implement these semantics as a library, rather than as a new dialect of Python.
+Python does not have a static type system, however, so to implement these semantics, we will use \verb|atlang|, an extensible type system for Python (being developed by the authors).
+By using \verb|atlang|, which leverages Python's quotations and reflection facilities, we can implement these semantics as a library, rather than as a new dialect of the language.
 
@@
 \begin{figure}[h]\begin{lstlisting}
-from atlib import fn, string_in
+from atlib import fn, stringin
 
 @fn
-def sanitize(s : string_in[r'.*']):
+def sanitize(s : stringin[r'.*']):
   return (s.replace(r'"', '&quot;') 
@@
 @fn
-def results_query(s : string_in[r'[^"]*']):
+def results_query(s : stringin[r'[^"]*']):
   return 'SELECT * FROM users WHERE name="' + s + '"'
@@
 @fn
-def results_div(s : string_in[r'[^<>]*']):
+def results_div(s : stringin[r'[^<>]*']):
   return '<div>Results for ' + s + '</div>'
@@
 @fn
-def main(db):
+def main():
   input = sanitize(user_input())
-  results = db.execute(results_query(input))
+  results = db_execute(results_query(input))
   return results_div(input) + format(results)
@@
 
-Figure \ref{fig:atexample} demonstrates the use of two type constructors, \verb|fn| and \verb|string_in|, both of which we have included in \verb|atlib|, the standard library for \verb|atlang|.
-The \verb|fn| type constructor can be used to annotate functions that should be statically checked by \verb|atlang|.
-The function \verb|sanitize| on lines 3-7 specifies one argument, \verb|s|, of type \lstinline{string_in[r'.*']}.\todo{fix sizes}
-
+\begin{figure}
+\begin{lstlisting}
+import re
+
+def sanitize(s):
+   return re.sub(r'"', re.sub(r'<', re.sub(r'>', 
+     s, '&gt;'), '&lt;'), '&quot;')
+
+def results_query(s): 
+  return 'SELECT * FROM users WHERE name="' + s + '"'
+
+def results_div(s):
+  return '<div>Results for ' + s + '</div>'
+
+def main():
+  input = sanitize(user_input())
+  results = db_execute(results_query(input))
+  return results_div(input) + format(results)
+\end{lstlisting}
+\vspace{-10px}
+\caption{The output of compilation of Figure \ref{fig:atexample} (at the terminal, \texttt{atc figure9.py}, or just-in-time).}
+\label{fig:atexample-out}
+\end{figure}
+
+\newcommand{\lstinlinep}[1]{\lstinline[language=Python,basicstyle=\ttfamily\small]{#1}}
+
+Figure \ref{fig:atexample} demonstrates the use of two type constructors, \verb|fn| and \verb|stringin|, corresponding to the two type constructors of $\lambda_{RS}$.
+We show these as being imported from \verb|atlib|, the standard library for \verb|atlang| (it benefits from no special support from the language itself).
+
+The \verb|fn| type constructor can be used to annotate functions that should be statically checked by \verb|atlang|.\footnote{Here, we use argument annotation syntax only available in versions 3.0+ of Python.
+Alternative syntax supporting Python 2.7+ can also be used, but we omit it here.} The function \verb|sanitize| on lines 3-7, for example,  specifies one argument, \verb|s|, of type \lstinlinep{stringin[r'.*']}.
+Here, the index is a regular expression, written using Python's \emph{raw string notation} as is conventional (in this particular instance, the \verb|r| is not strictly necessary).
 The sanitize function takes an arbitrary string and returns a string without
-double quotes or left and right brackets. In this example, we use HTML
-escape sequences.
-
-The main function receives user input and passes this input to a 
-sanitize function, which replaces all double quotes and brackets with HTML
-escape sequences.
-
-The result of applying sanitize to input is appended in two functions which
-construct a safe query (avoiding command injection) and safe HTML output (avoiding Cross-Site Scripting (XSS) attacks).
-The arguments to the result and output construction functions constitute \emph{specifications}.
-In the case of \textsf{results_query}, this specification ensures that user input is always interpreted
-as a string literal by the SQL server.
+double quotes or left and right brackets. Note that the return type need not be specified: \verb|atlang| uses a form of \emph{local type inference} \cite{Pierce:2000:LTI:345099.345100}.
+
+In this example, we use an HTML encoding so that the same sanitization function can be used to generate both SQL commands and HTML securely.
+The sanitized string is generated by invoking the \verb|replace| operator, which has the same semantics as $\textsf{rreplace}$ in $\lambda_{RS}$.
+Unlike in the core calculus, it is invoked like a method on \verb|s|.
+The regular expression determining the substrings to be replaced is given as the first argument (as in $\lambda_{RS}$, the only restriction here is that the regular expression must be specified statically.)
+
+The functions \verb|results_query| and \verb|results_div| construct a SQL query and an HTML snippet, respectively, by regular string concatenation.
+The argument type annotations serve as a check that sanitation was properly performed.
+In the case of \textsf{results_query}, this specification ensures that user input cannot prematurely be terminated.
 In the case of \textsf{results_div}, this specification ensures that user input does 
 not contain any HTML tags, which is a conservative but effective policy for preventing XSS attacks.
-
-Note that ${\sf input}$ does not actually meet these specifications without additional machinery.
-The type of ${\sf input}$ is quite large
-and does not actually equal the specified domains of the query or output construction
-methods. This mismatch is common -- in fact, nearly universal. Therefore, 
-our implementation includes a simple subtyping relation between regular expression
-types.
-
-This subtyping relation is justified theoretically by the fact that language inclusion is
-decidable; see \cite{fulton12} for a formal definition of the subtyping relation.
-Additionally, our extension remains composable because subtyping is defined on a type-by-type basis;
-see \cite{fulton13} for a discussion of subtyping in Atlang (referred to there as Ace).
-
-\subsection{Implementation of the Regular Expression Type}
+Note that the type of the surrounding string literals are determined by the type constructor of the function they appear in, \verb|fn| in both cases, which we assume simply chooses \lstinlinep{stringin[r'.*']} (an alternative strategy would be to use the most specific type for the literal, rather than the most general, but this choice is immaterial for this example).
+The addition operator here corresponds to the $\textsf{rconcat}$ operator in $\lambda_{RS}$.
+
+The \verb|main| function invokes the functions just described.
+It begins by passing  user input to \verb|sanitize|, then executing a database query and returning HTML based on this sanitized input.
+The helper functions \verb|user_input| and \verb|db_execute| are not shown but can be assumed standard.
+Importantly, had we mistakenly forgotten to call \verb|sanitize|, the function would not type check (in this case, it is obvious that we did, but lines 19 and 20 would in practice be separated more drastically in the code).
+Moreover, had \verb|sanitize| itself not been implemented correctly (e.g.
+we forgot to strip out quotation marks), then \verb|main| would also not typecheck either.
+
+One somewhat subtle issue here is that the return type of \verb|sanitize| is equivalent to \lstinlinep{stringin[r'[^"<>]*']}, which is a distinct type from the argument types to \verb|results_query| and \verb|results_div|.
+More specifically, however, it is a ``smaller'' type, in that it could be coerced to these argument types using an operator like $\textsf{rcoerce}$ in $\lambda_{RS}$.
+In our implementation, safe coercions are performed implicitly rather than explicitly.
+Because all regular strings are implemented as strings, this coercion induces no run-time change in representation.
+
+%The result of applying sanitize to input is appended in two functions which
+%construct a safe query (avoiding command injection) and safe HTML output (avoiding Cross-Site Scripting (XSS) attacks).
+%The arguments to the result and output construction functions constitute \emph{specifications}.
+
+%Note that ${\sf input}$ does not actually meet these specifications without additional machinery.
+%The type of ${\sf input}$ is quite large
+%and does not actually equal the specified domains of the query or output construction
+%methods. This mismatch is common -- in fact, nearly universal. Therefore, 
+%our implementation includes a simple subtyping relation between regular expression
+%types.
+
+%This subtyping relation is justified theoretically by the fact that language inclusion is
+%decidable; see \cite{fulton12} for a formal definition of the subtyping relation.
+%Additionally, our extension remains composable because subtyping is defined on a type-by-type basis;
+%see \cite{fulton13} for a discussion of subtyping in Atlang (referred to there as Ace).
+Figure \ref{fig:atexample-out} shows the output of typechecking and translating this code (this can occur either in batch mode at the terminal, generating a new file, or ``just-in-time'' at the first callsite of each function in the dynamically typed portion of the program, not shown).
+
+\subsection{Implementation}
 
@@
 \begin{lstlisting}
-class string_in(atlang.Type):
+class stringin(atlang.Type):
   def __init__(self, rx):
-    rx = rx_normalize(rx)
     atlang.Type.__init__(idx=rx)
@@
     right_t = ctx.syn(node.right)
-    if isinstance(left_t, string_in):
+    if isinstance(left_t, stringin):
       left_rx = left_t.idx
-      if isinstance(right_t, string_in):
+      if isinstance(right_t, stringin):
         right_rx = right_t.idx
-        return string_in[lconcat(left_rx, right_rx)]
+        return stringin[lconcat(left_rx, right_rx)]
     raise atlang.TypeError("...", node)
@@
     exp_t = ctx.syn(exp)
-    if not isinstance(exp_t, string_in):
+    if not isinstance(exp_t, stringin):
       raise atlang.TypeError("...", node)
     exp_rx = exp_t.idx
-    return string_in[lreplace(self.idx, rx, exp_rx)]
+    return stringin[lreplace(self.idx, rx, exp_rx)]
 
@@
 
-  def syn_Method_check(self, ctx, node):
-    [rx] = node.args
-    if not isinstance(rx, ast.Str):
-      raise atlang.TypeError("...", node)
-    return string_in[rx.s]
-
-  def trans_Method_check(self, ctx, node):
-    return astx.quote(
-      """__import__(string_in_helper);
-      string_in_helper.coerce(%0, %1)""",
-      astx.Str(s=other_t.idx),
-      astx.copy(node))
+  # check and strcase omitted
 
@@
 \vspace{-10px}
-\caption{Implementation of the \texttt{string\_in} type constructor in atlang.}
+\caption{Implementation of the \texttt{stringin} type constructor in atlang.}
 \label{fig:impl}
@@
 \label{fig:atimpl}\end{figure}
-
-We implemented a variation on the type system presented in this paper with two
-significant differences. first, we only support replacements where $s_2$ is the empty
-string. Therefore, our implementation respects the system presented in section 2
-only modulo the definition of \textsf{lreplace}. Seocnd, we use subtyping instead of
-\textsc{S-*-SafeCoerce}, which decreases inessential verbosity of the language.
-
-Atlang translates programs using type definitions, which may extend both the
-static and dynamic semantics of the language. New types are defined as Python
-classes; figure \ref{fig:impl} contains the source code of our implementation.
-
-The \textsf{string_in} type has an indexing regular expression \texttt{idx}.
-Our translation is defined by the \textsf{trans_} methods while the \textsf{syn_}
-methods define our type checker. Atlang defers type checking and translation
-to these methods whenever an expression of type \textsf{string_in} is encountered.
-
-\section{Related Work and Alternative Approaches}\label{related}
+The primary unit of extension in \verb|atlang| is the \emph{active type constructor}, rather than the abstract syntax as in other work on language extensibility.
+This allows us to implement the entire system as a library in Python and avoid needing to develop new tooling, and also makes it more difficult to create ambiguities between extensions.
+Active type constructors are subclasses of \verb|atlang.Type|, and types are instances of these classes.
+The methods of active type constructors control how typechecking and translation proceed for associated operations.
+In Figure \ref{fig:impl}, we show key portions of the implementation of the \verb|stringin| type used in the example above.
+Although a detailed description of the extension mechanism is beyond the scope of this work, we describe the intuitions behind the various methods below.
+
+The constructor, \verb|__init__| in Python, is called when a type is constructed. It simply stores the provided regular expression as the type index by calling the superclass.
+
+The method \verb|ana_Str| is called when a string literal is being checked against a regular string type, \lstinlinec{self}.
+It checks that the string is in the language of the regular expression provided as the type index, corresponding to rule \textsc{S-T-Stringin-I} in Section 2.
+The method \verb|trans_Str| is called after typechecking to produce a translation.
+Here, we just copy the original string literal -- regular strings are implemented as strings.
+
+The method \verb|syn_BinOp_Add| synthesizes a type for the string concatenation operation if both arguments are regular strings, consistent with rule \textsc{S-T-Concat}.
+The corresponding method \verb|trans_BinOp_Add| again simply translates the operation directly to string concatenation, consistent with our translation semantics.
+
+The method \verb|syn_Method_replace| synthesizes a type for the ``method-like'' operator \verb|replace|, seen used in our example.
+It ensures that the first argument is a statically known string, using Python's built-in \verb|ast| module, and leverages an implementation of \verb|lreplace|, which computes the appropriate regular expression for the string following replacement, again consistent with our description in Section 2.
+Translation proceeds by using the \verb|re| library built into Python, as can be seen in Figure \ref{fig:atexample-out}.
+
+Code for checked conversions and string decomposition is omitted, but is again consistent with our specification in the previous section.
+Safe coercions are controlled by the \verb|check_Coerce| function, which checks for a sublanguage relationship.
+Here, as in the other methods, failure is indicated by raising an \verb|atlang.TypeError| with an appropriate error message and location.
+
+Taken together, we see that the mechanics of extending \verb|atlang| with a new type constructor are fairly straightforward: we determine which syntactic forms the operators we specified in our core calculus should correspond to, then directly implement a decision procedure for type synthesis (or, in the case of literal forms, type analysis) in Python.
+The \verb|atlang| compiler invokes this logic when working with regular strings.
+The result is a library-based embedding of our semantics into Python.
+
+% We implemented a variation on the type system presented in this paper with two
+% significant differences. first, we only support replacements where $s_2$ is the empty
+% string. Therefore, our implementation respects the system presented in section 2
+% only modulo the definition of \textsf{lreplace}. Seocnd, we use subtyping instead of
+% \textsc{S-*-SafeCoerce}, which decreases inessential verbosity of the language.
+
+% Atlang translates programs using type definitions, which may extend both the
+% static and dynamic semantics of the language. New types are defined as Python
+% classes; figure \ref{fig:impl} contains the source code of our implementation.
+
+% The \textsf{stringin} type has an indexing regular expression \texttt{idx}.
+% Our translation is defined by the \textsf{trans_} methods while the \textsf{syn_}
+% methods define our type checker. Atlang defers type checking and translation
+% to these methods whenever an expression of type \textsf{stringin} is encountered.
+
+\section{Related Work}\label{related}
 
@@
 %\hline
-%${\tt string\_in}[r]$ & $\fvar{stringin}[\xlA{r}]$ & \verb|string_in[|\xA{r}\verb|]|\\
+%${\tt string\_in}[r]$ & $\fvar{stringin}[\xlA{r}]$ & \verb|stringin[|\xA{r}\verb|]|\\
 %\\

### 2014-08-21 sanitation-psp14/sanitation-psp14-short.tex
updates per JA

@@
 In this paper, we introduce \emph{regular string types}, which classify strings known statically to be in a specified regular language.
-These types come equipped with common operations like concatenation, substitution and coercion, so they  can be used to implement, in essentially a conventional manner, the portions of a web application or web application framework that must directly construct command strings.
+These types come equipped with common operations like concatenation, substitution and coercion, so they  can be used to implement, in a conventional manner, the portions of a web application or application framework that must directly construct command strings.
 Simple type annotations at key interfaces can be used to statically verify that sanitization has been performed correctly without introducing redundant run-time checks.
-We specify this type system as a minimal typed lambda calculus, $\lambda_{RS}$.
+We specify this type system in a minimal typed lambda calculus, $\lambda_{RS}$.
 
@@
 Instead, we advocate for extensible type systems: new type system fragments like this should be implemented as libraries atop a mechanism that guarantees that they can be safely composed.
-We support this by 1) specifying a translation from $\lambda_{RS}$ to a language containing only strings and regular expressions, then, taking Python as such a language, 2) implementing the type system together with the translation as a library using \texttt{atlang}, an extensible static type system for Python (being developed by the authors).% represents the most sophisticated application of this library to date.
+We support this by 1) specifying a translation from $\lambda_{RS}$ to a language fragment containing only standard strings and regular expressions, then, taking Python as a language with these constructs, 2) implementing the type system together with the translation as a library using \texttt{atlang}, an extensible static type system for Python being developed by the authors.% represents the most sophisticated application of this library to date.
 
@@
 Alternatively, it might HTML-encode special characters, which would head off both SQL injection attacks and cross-site scripting attacks.
-Guaranteeing that user input has already been sanitized  before it is used to construct a command is challenging, because  sanitation is often performed in a different part of the code than the place where commands are ultimately constructed.% Note, for example, that \verb|sanitize| is not idempotent, so it should only be called once.
+Guaranteeing that user input has already been sanitized  before it is used to construct a command is challenging, because  sanitization is often performed in a different part of a program than where commands are ultimately constructed.% Note, for example, that \verb|sanitize| is not idempotent, so it should only be called once.
 
@@
 By leveraging closure and decidability properties of regular languages, the type system tracks the language of a string through uses of a number of operations, including \emph{replacement} of substrings matching a given pattern.
-This makes it possible to implement sanitation functions like the one just described in an essentially conventional manner.
+This makes it possible to implement sanitation functions like the one just described in a conventional manner.
 The result is a system where the fact that a string has been {correctly} sanitized is manifest in its type.
@@
 Waiting for a language designer to build this feature in is unsatisfying in practice.
-Moreover, we also face a ``chicken-and-egg problem'': justifying its inclusion into a commonly used language benefits from empirical demonstrations that it is useful, but this is difficult to do if developers have no way to use it in practice.
-As such, we take the position that a better path forward for the community is to work within a programming language where such type system fragments can be introduced modularly and orthogonally, as libraries.
+Moreover, we also face a ``chicken-and-egg problem'': justifying its inclusion into a commonly used language benefits from empirical case studies, but these are  difficult to conduct if developers have no way to use the abstraction in real-world code.
+We take the position that a better path forward for the community is to work within a programming language where such type system fragments can be introduced modularly and orthogonally, as libraries.
 
@@
 \ref{atlang}, we show how to implement the type system fragment we have specified  using \texttt{atlang}, an extensible static type system implemented as a library inside Python.
-\texttt{atlang} leverages local type inference to control the semantics of literal forms, so reguar string types can be introduced using string literals without any run-time overhead.
+\texttt{atlang} leverages local type inference to control the semantics of literal forms, so regular string types can be introduced using string literals without any run-time overhead.
 Coercions that are known to be  safe due to a sublanguage relationship are performed implicitly, also without run-time overhead.
@@
 \item Section \ref{sec:p} describes a simple target language, $\lambdap$, with a minimal regular expression library. In Section \ref{atlang}, we will take Python to be such a language.
-\item Section \ref{sec:tr} describes the translation from $\lambdas$ to $\lambdap$ and ensures the correcness result for \ref{sec:rs} is preserved under translation. 
+\item Section \ref{sec:tr} describes the translation from $\lambdas$ to $\lambdap$ and ensures the correctness result for \ref{sec:rs} is preserved under translation. 
 \end{itemize}
@@
 
-Fortunately, regular expression derivatives \cite{bowzer} convienantly capture this intuition.
+Fortunately, regular expression derivatives \cite{bowzer} conveniently capture this intuition.
 The regular expression recognizing any one-character prefix of the strings in $\lang{r}$ is easily defined.
@@
 
-The method \verb|ana_Str| is called when a string literal is being checked against a regular string type, \lstinlinec{self}.
+When a string literal is being checked against a regular string type, the method \verb|ana_Str| is called.
 It checks that the string is in the language of the regular expression provided as the type index, corresponding to rule \textsc{S-T-Stringin-I} in Section 2.

### 2014-08-21 sanitation-psp14/sanitation-psp14-short.tex
JA sec 2

@@
 
-The $\lambdas$ language supports two forms of coercion. Safe coercion only allows coercsion between 
+The $\lambdas$ language supports two forms of coercion. Safe coercion only allows coercion between 
 strings types which are guaranteed to be safe.
@@
     expoiting the decidability of language inclusion, while
-\textsc{S-T-Check} allows casts between strings that cannot be guaranteed at compile time, but inserts a runtime check a runtime check.
+\textsc{S-T-Check} allows casts between strings that cannot be guaranteed at compile time, but inserts a runtime check.
+
+An alternative to coercion is subtyping for constrained string types.
+For simplicity, we do not include this relation; however, subtyping for constrained strings is present in \cite{fulton13} and that treatment extends to this system.
 
@@
 On strings, the string substitution operator is familiar, and roughly analogous to \lstinline{str_replace} in PHP and \lstinline{String.replace} in Java.
-The definition of \textsc{S-T-Case} requires a definition of replacement for regular expressions (or languages) corresponding to string replacement.
+The definition of \textsc{S-T-Replace} requires a definition of replacement for regular expressions (or languages) corresponding to string replacement.
 In this section, we define this operation.
@@
 The function $\lsubst{r}{s_1}{s_2}$ replaces every substring of $s_1$ matching $r$ with $s_2$.
-In an analogous manner, the function $\lreplace{r}{r_1}{r_2}$ replaces every sublanguages of $r_1$ matching $r$ with $r_2$.
-If an intuiition for this operation is not clear, it may be helpful to think in terms replacing sub-automata.
+In an analogous manner, the function $\lreplace{r}{r_1}{r_2}$ replaces every sublanguage of $r_1$ matching $r$ with $r_2$.
+If the intuition for this operation is not clear, it may be helpful to think in terms of replacing sub-automata.
 
@@
   a language $r'$ containing all strings of $r_1$ except that any substring $s_{pre} s s_{post} \in \lang{r_1}$ where $s \in \lang{r}$
-  is replaced by the set of strings $s_{pre} s_2 s_{post}$ for all $s_2 \in \lang{r_2}$ (the prefix and postfix positions may be empty).
+  is replaced by the set of strings $s_{pre} s_2 s_{post}$ for all $s_2 \in \lang{r_2}$ (the prefix and postfix positions may be empty). This procedure saturates the string.
 \end{defn}
@@
 \end{proof}
-\begin{lem}
-If $\tctx \vdash \iota : \regex$ then $\iota \treduces \rx{r}$ such that $r$ is a well-formed regular expression. 
-\end{lem}
 
@@
   If $\sctx \vdash e : \sigma$ 
-  then $e \sreduces e'$ and $\sctx \vdash e' : \stringin{r}$.
+  then $e \sreduces v$ and $\sctx \vdash v : \sigma$.
 \end{thm}
@@
 note the multiple approaches to proving soundness for non-terminating languages with natural semantics \cite{coinductionoo, stepindexing, coinduction}.
+
+\begin{lem}
+If $\tctx \vdash \iota : \regex$ then $\iota \treduces \rx{r}$ such that $r$ is a well-formed regular expression. 
+\end{lem}
 

### 2014-08-21 sanitation-psp14/sanitation-psp14-short.tex
Apocalypse

@@
 \begin{figure}[h]
+\small
   \begin{grammar}
@@
       $\strin{s}$ \hfill $s \in \Sigma^{*}$ \alt
-      $\rsconcat{e}{e}$ \alt
-      $\strcase{e}{e}{x,y.e}$ \alt
+      $\rsconcat{e}{e}$ | $\strcase{e}{e}{x,y.e}$ \alt
       $\sisubst{r}{e}{e}$ \alt
-      $\rcoerce{r}{e}$ \alt
-      $\rcheck{r}{e}{x.e}{e}$
-
-<$v$> ::= $\lambda x . e$ \hfill source values \alt
-          $\strin{s}$ \hfill $s \in \Sigma^{*}$
-
+      $\rcoerce{r}{e}$ | $\rcheck{r}{e}{x.e}{e}$
+
+      <$v$> ::= $\lambda x . e$ | $\strin{e}$ | $s$ \hfill source values 
 \caption{Syntax of $\lambda_{RS}$.}
@@
 Intuitively, this rule ``peels off" the first character, and then performs a specified operation on the first character and the remaining string.
-In terms of strings, this is essentially the ubiquitious substring operation.
+
+In terms of strings, \textsc{S-T-Case} is essentially the ubiquitious substring operation.
 
@@
 The function $\lhead{r}$ is defined in terms of the structure of $r$:
-\begin{itemize}
+\begin{itemize}[noitemsep]
 \item $\lhead{a r'} = a$ where $a \in \Sigma$
@@
 
-An alternative to coercion is subtyping for constrained string types.
-For simplicity, we do not include this relation; however, subtyping for constrained strings is present in \cite{fulton13} and that treatment extends to this system.
+An alternative to coercion is subtyping for constrained string types, which applies coercion implicitly. 
+In practice, subtyping is crucial to the usability of our system due to the nature of the replacement operator.
+For simplicity, we do not include this relation; however, subtyping for constrained strings is present in \cite{fulton12,fulton13} and that treatment extends to this system.
+The implementation discussed in Section \ref{atlang} provides subtyping between constrained string types.
 
@@
 \begin{figure}[t]
+\small
   \begin{grammar}
@@
   $\rx{r}$ \alt
-  $\tconcat{\iota}{\iota}$ \alt
-  $\pstrcase{\iota}{\iota}{x,y.\iota}$ \alt
-  $\preplace{\iota}{\iota}{\iota}$ \alt
-  $\tcheck{\iota}{\iota}{\iota}{\iota}$
-
-<$\dot{v}$> ::= $\lambda x . \iota$ \hfill target values \alt
-                $\tstr{s}$ \alt
-                $\rx{r}$
+  $\tconcat{\iota}{\iota}$ | $\pstrcase{\iota}{\iota}{x,y.\iota}$ \alt
+  $\preplace{\iota}{\iota}{\iota}$ | $\tcheck{\iota}{\iota}{\iota}{\iota}$
+
+  <$\dot{v}$> ::= $\lambda x . \iota$ | \tstr{s} | \rx{r} \hfill target values
 
@@
 
-In practice, the type of a replacement rarely matches a specification.
-Therefore, it is convienant in an implementation to always insert the appropriate
-safe coercion. This way, violations of the sanitation protocol manifest as type errors,
-but the definition of an adequate replacement is not arduous.
-Alternatively, this policy may 
-be codified in the type system itself using subtyping \cite{fulton12}.
-
 The translation is defined by the rules \textsc{Tr-} in Figure~\ref{fig:tr}.
@@
 \begin{figure}[h]
+\small
   $\fbox{\inferrule{}{\trden{\sigma}=\tau}}$
@@
 
-Incorporating regular expressions into the type system is not novel.
-The XDuce system \cite{HosoyaVouillonPierce2000ICFP,HosoyaPierce2002} checks XML documents against schema using regular expressions.
+% HosoyaPierce2002? 
+Incorporating regular expressions into the type system is not novel. The XDuce system \cite{HosoyaVouillonPierce2000ICFP} checks XML documents against schema using regular expressions. 
 Similarly, XHaskell \cite{xhaskell} focuses on XML documents.
@@
 
-We believe that this type system extension serves as a useful basis for web-oriented static analysis.
-The use sites of arbitrary strings are usually constrained to a few sections of the application. Assuming
-regular expressions are used for input sanitation, annotations at these use sites could be sufficient input to a sound and complete static analysis.
-
-We also believe that extensible programming languages are a promising approach toward incorporating security
-and privacy into programming languages. The Software Industry moves quickly, and developing new languages for 
-the particular security and privacy concerns of each new domain is likely a losing strategy. Conversely, extensible
-languages might provide developers with a way to incorporate domain-specific security and privacy analyses into their
-daily development activities. In future work, we hope to bolster this hypothesis by extending our approach in this paper
-to more realistic settings and by developing new type extensions addressing security and privacy concerns.
-
+We believe that this type system extension serves as a useful basis for web-oriented static analysis;
+frameworks and regular expression libraries could be annotated, along with use-sites.
+%The use sites of arbitrary strings are usually constrained to a few sections of the application. Assuming
+%regular expressions are used for input sanitation, annotations at these use sites could be sufficient input to a sound and complete static analysis.
+%
+We also believe that extensible programming languages are a promising approach toward incorporating other security
+and privacy analyses into programming languages.
+%languages might provide developers with a way to incorporate domain-specific security and privacy analyses into their
+%daily development activities. In future work, we hope to bolster this hypothesis by extending our approach in this paper
+%to more realistic settings and by developing new type extensions addressing security and privacy concerns.
+%
 \section{Conclusion}

### 2014-08-21 sanitation-psp14/sanitation-psp14-short.tex
updated printout and caught a typo. True Apoc.

@@
 \begin{proof}
-The proof proceeds by induction on the typing relation for $e$ and an appropriate choich of $\iota$; in each case,
+The proof proceeds by induction on the typing relation for $e$ and an appropriate choice of $\iota$; in each case,
 the choice is the syntactic form in $\lambdap$ corresponding to the form under consideration (e.g. choose {\sf replace} when considering {\sf sreplace}).

### 2014-08-21 sanitation-psp14/sanitation-psp14-short.tex
Added a TXT containing the submitted abstract, as well as a document for any errors we find before hearing about from reviewers. Also made   some minor changes.

@@
 \begin{abstract}
-%Evidence suggests that programmers are reluctant to adopt new languages to
 Web applications must ultimately command systems like web browsers and database engines using strings.
 Strings derived from improperly sanitized user input  can thus be a vector for command injection attacks.
-%Structured frameworks for constructing commands centralizes control over string sanitization, but verifying that the framework itself sanitizes user input correctly can still be tricky.
 In this paper, we introduce \emph{regular string types}, which classify strings known statically to be in a specified regular language.
@@
 Instead, we advocate for extensible type systems: new type system fragments like this should be implemented as libraries atop a mechanism that guarantees that they can be safely composed.
-We support this by 1) specifying a translation from $\lambda_{RS}$ to a language fragment containing only standard strings and regular expressions, then, taking Python as a language with these constructs, 2) implementing the type system together with the translation as a library using \texttt{atlang}, an extensible static type system for Python being developed by the authors.% represents the most sophisticated application of this library to date.
-
-% Security-oriented programming languages do not enjoy wide adoption.
-Conversely, library and framework-based approaches to security are pervasive.
-Unfortunately, libraries and frameworks are only effective solutions if they are implemented and used properly.
-In this paper, we propose that extensible programming languages provide a promising mechanism for increasing the amount of trust that can be place in libraries and frameworks.
-To demonstrate this approach, we present a conservative extension to the simply-typed lambda calculus for checking the correctness of input sanitation algorithms.
-We also demonstrate how this language's security guarantees are preserved under translation into an underlying language containing a regular expression library.
-We believe this approach -- complementing existing techniques with light-weight, composable type-based analyses -- constitutes a promising compromise between usability, potential for industry adoption, and theoretically grounded safety guarantees.
+We support this with two contributions.
+First, we specify a translation from $\lambda_{RS}$ to a language fragment containing only standard strings and regular expressions.
+Second, taking Python as a language with these constructs, we implement the type system together with the translation as a library using \texttt{atlang}, an extensible static type system for Python being developed by the authors.
 \end{abstract}
@@
 By leveraging closure and decidability properties of regular languages, the type system tracks the language of a string through uses of a number of operations, including \emph{replacement} of substrings matching a given pattern.
-This makes it possible to implement sanitation functions like the one just described in a conventional manner.
+This makes it possible to implement sanitation functions -- like the one just described -- in a conventional manner.
 The result is a system where the fact that a string has been {correctly} sanitized is manifest in its type.
@@
 
-The \textsc{S-T-Concat} rule, which is defined in terms of these operations, relates the result of ``peeling off" the first character of a string to regular expression derivatives.
+The \textsc{S-T-Case-Concat} rule, which is defined in terms of these operations, relates the result of ``peeling off" the first character of a string to regular expression derivatives.
 

### 2014-08-21 sanitation-psp14/sanitation-psp14-short.tex
updated

--- 2014-08-21 sanitation-psp14/sanitation-psp14-short.tex
@@
 \begin{figure}[h]
-\small
 \begin{grammar}
@@
       $x$ \hfill source terms \alt 
-      $\lambda x . e$ \alt
+      $\lambda x . e$ e\alt
       $e(e)$ \alt
       $\strin{s}$ \hfill $s \in \Sigma^{*}$ \alt
-      $\rsconcat{e}{e}$ | $\strcase{e}{e}{x,y.e}$ \alt
+      $\rsconcat{e}{e}$ \alt
+      $\strcase{e}{e}{x,y.e}$ \alt
       $\sisubst{r}{e}{e}$ \alt
-      $\rcoerce{r}{e}$ | $\rcheck{r}{e}{x.e}{e}$
-
-      <$v$> ::= $\lambda x . e$ | $\strin{e}$ | $s$ \hfill source values 
+      $\rcoerce{r}{e}$ \alt
+      $\rcheck{r}{e}{x.e}{e}$
+
+<$v$> ::= $\lambda x . e$ \hfill source values \alt
+          $\strin{s}$ \hfill $s \in \Sigma^{*}$
+
 \caption{Syntax of $\lambda_{RS}$.}
@@
 Intuitively, this rule ``peels off" the first character, and then performs a specified operation on the first character and the remaining string.
-
-In terms of strings, \textsc{S-T-Case} is essentially the ubiquitious substring operation.
+In terms of strings, this is essentially the ubiquitious substring operation.
 
@@
 The function $\lhead{r}$ is defined in terms of the structure of $r$:
-\begin{itemize}[noitemsep]
+\begin{itemize}
 \item $\lhead{a r'} = a$ where $a \in \Sigma$
@@
 
-An alternative to coercion is subtyping for constrained string types, which applies coercion implicitly. 
-In practice, subtyping is crucial to the usability of our system due to the nature of the replacement operator.
-For simplicity, we do not include this relation; however, subtyping for constrained strings is present in \cite{fulton12,fulton13} and that treatment extends to this system.
-The implementation discussed in Section \ref{atlang} provides subtyping between constrained string types.
+An alternative to coercion is subtyping for constrained string types.
+For simplicity, we do not include this relation; however, subtyping for constrained strings is present in \cite{fulton13} and that treatment extends to this system.
 
@@
 \begin{figure}[t]
-\small
 \begin{grammar}
@@
   $\rx{r}$ \alt
-  $\tconcat{\iota}{\iota}$ | $\pstrcase{\iota}{\iota}{x,y.\iota}$ \alt
-  $\preplace{\iota}{\iota}{\iota}$ | $\tcheck{\iota}{\iota}{\iota}{\iota}$
-
-  <$\dot{v}$> ::= $\lambda x . \iota$ | \tstr{s} | \rx{r} \hfill target values
+  $\tconcat{\iota}{\iota}$ \alt
+  $\pstrcase{\iota}{\iota}{x,y.\iota}$ \alt
+  $\preplace{\iota}{\iota}{\iota}$ \alt
+  $\tcheck{\iota}{\iota}{\iota}{\iota}$
+
+<$\dot{v}$> ::= $\lambda x . \iota$ \hfill target values \alt
+                $\tstr{s}$ \alt
+                $\rx{r}$
 
@@
 If the safety of coercion is not manifest in the types, then a check is inserted.
+
+In practice, the type of a replacement rarely matches a specification.
+Therefore, it is convienant in an implementation to always insert the appropriate
+safe coercion. This way, violations of the sanitation protocol manifest as type errors,
+but the definition of an adequate replacement is not arduous.
+Alternatively, this policy may 
+be codified in the type system itself using subtyping \cite{fulton12}.
 
@@
 
-\begin{figure}[h]\begin{lstlisting}
+\begin{figure}[h]\small\begin{lstlisting}
 from atlib import fn, stringin
@@
 
-% HosoyaPierce2002? 
-Incorporating regular expressions into the type system is not novel. The XDuce system \cite{HosoyaVouillonPierce2000ICFP} checks XML documents against schema using regular expressions. 
+Incorporating regular expressions into the type system is not novel.
+The XDuce system \cite{HosoyaVouillonPierce2000ICFP,HosoyaPierce2002} checks XML documents against schema using regular expressions.
 Similarly, XHaskell \cite{xhaskell} focuses on XML documents.
@@
 
-We believe that this type system extension serves as a useful basis for web-oriented static analysis;
-frameworks and regular expression libraries could be annotated, along with use-sites.
-%The use sites of arbitrary strings are usually constrained to a few sections of the application. Assuming
-%regular expressions are used for input sanitation, annotations at these use sites could be sufficient input to a sound and complete static analysis.
-%
-We also believe that extensible programming languages are a promising approach toward incorporating other security
-and privacy analyses into programming languages.
-%languages might provide developers with a way to incorporate domain-specific security and privacy analyses into their
-%daily development activities. In future work, we hope to bolster this hypothesis by extending our approach in this paper
-%to more realistic settings and by developing new type extensions addressing security and privacy concerns.
-%
+We believe that this type system extension serves as a useful basis for web-oriented static analysis.
+The use sites of arbitrary strings are usually constrained to a few sections of the application. Assuming
+regular expressions are used for input sanitation, annotations at these use sites could be sufficient input to a sound and complete static analysis.
+
+We also believe that extensible programming languages are a promising approach toward incorporating security
+and privacy into programming languages. The Software Industry moves quickly, and developing new languages for 
+the particular security and privacy concerns of each new domain is likely a losing strategy. Conversely, extensible
+languages might provide developers with a way to incorporate domain-specific security and privacy analyses into their
+daily development activities. In future work, we hope to bolster this hypothesis by extending our approach in this paper
+to more realistic settings and by developing new type extensions addressing security and privacy concerns.
+
 \section{Conclusion}

### 2014-08-27 thesis-proposal/acc.tex

@@
 }
+\newpage
 \section{Active Code Completion}\label{acc}
+\begin{figure}[h]\label{color}
+\begin{center}
+\includegraphics[width=40pc]{color_palette.png}\end{center}
+\caption{(a) An example code completion palette associated with the \texttt{Color} class. (b) The elaboration generated by this palette.}
+\label{colorpalette}
+\end{figure}
+
 An abstraction might benefit from support from a variety of tools beyond the compiler.
 For example, software developers today make heavy use of the code completion support found in modern source code editors  \cite{murphy_how_2006}.
-Editors for object-oriented languages provide code completion in the form of a floating menu containing  contextually-relevant variables, assignables, fields, methods, types and other code snippets.
+Code completion typically takes the form of a floating menu containing  contextually-relevant variables, assignables, fields, methods, types and other code snippets.
+When an item in the menu is selected, code is inserted immediately, without further input from the developer.
 By navigating and selecting from this menu, developers are able to avoid many common spelling and logic errors, eliminate unnecessary keystrokes and explore unfamiliar APIs without incurring the mental overhead associated with switching to an external documentation tool.
@@
 
-In all such systems, the code completion interface has remained primarily menu-based.
-When an item in the menu is selected, code is inserted immediately, without further input from the developer.
 These systems are difficult to extend: a fixed strategy determines the completions that are available, so library providers cannot directly specify new domain-specific or contextually-relevant logic or provide assistance beyond that which a menu can provide.
@@
 \end{lstlisting}
-
-\begin{figure*}\label{color}
-\begin{center}
-\includegraphics[width=40pc]{color_palette.png}\end{center}
-\caption{(a) An example code completion palette associated with the \texttt{Color} class. (b) The elaboration generated by this palette.}
-\label{colorpalette}
-\end{figure*}
 

### 2014-08-27 thesis-proposal/background.tex

@@
 Thus, macros represent at best a partial solution: they can be used to automate code generation, but not to globally extend the syntax or static guarantees of a language.
-It can also be difficult to reason about the semantics of code when any number of enclosing macros may be manipulating it, and to build tools that operate robustly in their presence.
+It can also be difficult to reason about the semantics of code when any number of enclosing macros may be manipulating it (particularly if they can change its type, as in some static macro systems), and to build tools that operate robustly in their presence.
 
@@
 For example, a library that defines a literal syntax for HTML would conflict with another that defines a literal syntax for XML because they define differing semantics for some of the same syntactic forms.
-If SugarJ was used by two different regular expression engines to provide literal syntax for regular expression patterns, there could easily be conflicts at link-time because both will introduce many of the same notations but back them with differing implementations.
+If SugarJ was used by two different regular expression engines to provide literal syntax for regular expression patterns, there could easily be conflicts at link-time because both will introduce many of the same syntax but back them with differing implementations.
 And again, it is difficult to predict what an unfamiliar piece of syntax desugars into, leading to difficulties reading and reasoning about code.

### 2014-08-27 thesis-proposal/parsing.tex

@@
 % !TEX root = omar-thesis-proposal.tex
-\section{Type-Specific Languages}\label{aparsing}
-General-purpose constructs are, semantically, quite flexible. 
+\section{Tycon-Specific Languages (TSLs)}\label{aparsing}
+Many abstractions can be seen, semantically, as modes of use of general-purpose constructs like sums and products. 
 %By using a general-purpose construct to define a data structure, one immediately benefits from a body of useful operators, established reasoning principles, well-optimized implementations and tool support.
-For example, lists can be defined using a more general mechanism for defining polymorphic recursive sum types by observing that a list can either be empty, or be broken down into a \emph{head} element and a \emph{tail}, another list.
-In an ML-like language, this would be written:
+For example, lists can be defined using a more general mechanism for defining polymorphic recursive sum types by observing that a list can either be empty, or be broken down into a product of the \emph{head} element and the \emph{tail}, another list.
+In an ML-like language, this would be written as a datatype:
 \begin{lstlisting}[numbers=none]
@@
 \end{lstlisting}
-By defining abstractions in terms of existing general purpose constructs, we immediately know how to reason about them (here, by structural induction) and examine them (by pattern matching).
+By defining type constructors like \li{list} in terms of these general purpose semantic mechanisms, we immediately know how to reason about them (here, by structural induction) and examine them (by pattern matching).
 They are already well optimized by compilers and benefit from general-purpose editor support as well.
 %The programmer only needs to provide an encoding of the structure of lists; the semantics and implementation are filled in by the general-purpose abstraction mechanism.
-While this is all quite useful, the associated general-purpose syntax is often less flexible.
+While this is all quite useful, the associated general-purpose concrete syntax is often less than ideal.
 For example, few would claim that writing a list of numbers as a sequence of \li{Cons} cells is convenient:
@@
 
-Because lists are a common data structure, many languages include \emph{literal notation} for introducing them, e.g.
+Because lists are a common data structure, many languages build in \emph{literal syntax} for introducing them, e.g.
 \li{[1, 2, 3, 4]}.
-This notation is semantically equivalent to the general-purpose notation shown above, but brings cognitive benefits by drawing attention to the content of the list, rather than the nature of the encoding.
+This syntax is semantically equivalent to the general-purpose syntax shown above, but brings cognitive benefits by drawing attention to the content of the list, rather than the nature of the encoding.
 Using terminology from Green's cognitive dimensions of notations \cite{green1996usability}, it is more \emph{terse}, \emph{visible} and \emph{maps more closely} to the intuitive notion of a list.
 Stoy, in discussing the value of good notation, writes \cite{stoy1977denotational}:
-\begin{quote}A good notation thus conceals much of the inner workings behind suitable abbreviations, while allowing us to consider it in more detail if we require: matrix and tensor notations provide further good examples of this.
-It may be summed up in the saying: ``A notation is important for what it leaves out.''\end{quote}
+\begin{quote}\emph{A good notation thus conceals much of the inner workings behind suitable abbreviations, while allowing us to consider it in more detail if we require: matrix and tensor notations provide further good examples of this.
+It may be summed up in the saying: ``A notation is important for what it leaves out.''}\end{quote}
 
-Although list, number and string literals are nearly ubiquitous features of modern languages, some languages also  provide specialized notation for other common data structures (like maps, sets, vectors and matrices), data formats (like XML and JSON), query languages (like regular expressions and SQL), markup languages (like HTML and Markdown) and many other types of data.
+List, number and string literals are nearly ubiquitous features of modern languages.
+Some languages additionally  build in  specialized notation for other common data structures (like maps, sets, vectors and matrices), data formats (like XML and JSON), query languages (like regular expressions and SQL), markup languages (like HTML and Markdown) and many other types of data.
 For example, a language with built-in notation for HTML and SQL, supporting type-safe \emph{splicing} via curly braces, might define:
@@
 \end{lstlisting}
-as shorthand for:
+as more concise and natural shorthand for:
 \begin{lstlisting}
@@
 
-When a specialized notation like this is not available, but the equivalent general-purpose notation is too cognitively demanding for comfort, developers typically turn to run-time mechanisms to make constructing data structures more convenient.
+When a specialized notation like this is not available, but this equivalent general-purpose notation, which in essence captures only the abstract syntax of languages like HTML and SQL, is unsatisfying, developers typically turn to run-time mechanisms to make constructing data structures more convenient.
 Among the most common strategies across language paradigms in these situations is to simply use a string representation that is parsed at run-time.% Developers across language paradigms frequently write examples like the above as:
@@
 
-Though recovering much of the notational convenience of the literal version, it is still more awkward to write, requiring explicit conversions to and from structured representations (\li{parse_html} and \li{to_string}, respectively) and escaping when the syntax of the language interferes with the syntax of string literals (line 2).
+Though recovering some of the notational convenience of the literal version, it is still more awkward, requiring explicit conversions to and from structured representations (\li{parse_html} and \li{to_string}, respectively) and escaping when the syntax of the language interferes with the syntax of string literals (line 2).
 Code like this also causes a number of problems beyond cognitive load.
@@
 
-Today, supporting new literal notations within an existing language requires the cooperation of the language designer.
+Today, supporting new literal syntax within an existing programming language requires the cooperation of the language designer.
 This is primarily because, with conventional parsing strategies, not all notations can unambiguously coexist, so a designer is needed to make choices about which syntactic forms are available and what their semantics should be.
@@
 In most existing related work, the burden of resolving ambiguities falls to clients who have the misfortune of importing conflicting extensions.
-For example, SugarJ \cite{erdweg2011sugarj} and other extensible languages generated by Sugar* \cite{erdweg2013framework} allow providers to extend the base syntax of the host language with new forms, like set and map literals.
+For example, SugarJ \cite{erdweg2011sugarj} and other extensible languages generated by Sugar* \cite{erdweg2013framework} allow providers to nearly arbitrarily extend the base syntax of the host language with new forms, like set and map literals.
 These new forms are imported transitively throughout a program.
@@
 In this work, we will describe an alternative parsing strategy that avoids these problems by shifting responsibility for parsing certain \emph{generic literal forms} into the typechecker.
-The typechecker, in turn, defers responsibility to user-defined types, by treating the body of the literal as a term of the   \emph{type-specific language (TSL)} associated with the type it is being checked against.
+The typechecker, in turn, defers responsibility to user-defined types, by treating the body of the literal as a term of the   \emph{type- or type-constructor-specific language (TSL)} associated with the type it is being checked against.
 The TSL is responsible for rewriting this term to ultimately use only general-purpose notation.
@@
 \end{lstlisting}
-\vspace{-8px}
+%\vspace{-8px}
 \caption{Wyvern Example with Multiple TSLs}
@@
 \subsection{Wyvern}
-We develop our work as a variant of a new programming language being developed by our group called Wyvern \cite{Nistor:2013:WST:2489828.2489830}.
+We specify our work as a variant of a new programming language being developed by our group called Wyvern \cite{Nistor:2013:WST:2489828.2489830}.
 To allow us to focus on the essence of our proposal, the variant of Wyvern we will describe in this thesis is simpler than the variant previously described: it is purely functional (there are no effects other than non-termination) and it does not enforce a uniform access principle for objects (fields can be accessed directly).
@@
 One can refer to the version of the language described in this thesis as \emph{TSL Wyvern}.
-TSL Wyvern has a layout-sensitive syntax, for reasons we will discuss.
+TSL Wyvern has a layout-sensitive concrete syntax, for reasons we will discuss.
 
@@
 The variable \li{imageBase} is annotated with its type, \li{URL}.
-This is an {object type} declaring several fields representing the components of a URL: its protocol, domain name, port, path and so on (not shown).
+This is a named {object type}\footnote{In our example we do not make use of parameterized types (i.e.
+\texttt{list}), so types and type constructors are indistinguishable, and we will simply use ``types'' for concision.} declaring several fields representing the components of a URL: its protocol, domain name, port, path and so on (not shown).
 We could have created a value of type \li{URL} using general-purpose notation:
-\begin{lstlisting}
+\begin{lstlisting}[numbers=none]
 let imageBase : URL = new
@@
 The type annotation on \li{imageBase} implies that this literal's \emph{expected type} is \li{URL}, so the {body} of the literal (the characters between the angle brackets, in blue) will be governed by the \li{URL} TSL during the typechecking phase.
-This TSL will parse the body ({at compile-time}) to produce a Wyvern abstract syntax tree (AST) that explicitly instantiates a new object of type \li{URL} using general-purpose notation as if the above had been written directly.
+This TSL specifies how to parse the body  to produce a Wyvern abstract syntax tree (AST) that explicitly instantiates a new object of type \li{URL} using general-purpose notation as if the above had been written directly.
 We will return to how this works shortly.
@@
 The TSL parses code between percent signs  as a Wyvern expression, using its abstract syntax tree (AST) to construct an AST for the expression as a whole.
-A string-based representation of the URL is never used at run-time.
+A string-based representation of the URL is never constructed.
 Note that the delimiters used to go from Wyvern to a TSL are controlled by Wyvern  while the TSL controls how to return to Wyvern.
@@
 We could again use Wyvern's general-purpose introductory form for case types, e.g.
-\li{HTML.BodyElement((attrs, child))} ({unlike in ML, in Wyvern we must explicitly qualify constructors with the case type they are part of when they are used}.
-This is largely to make our formal semantics simpler and for clarity of presentation.) But, as discussed above, using this syntax can be inconvenient and  cognitively demanding.
+\li{BodyElement((attrs, child))}.
+But, as discussed above, using this syntax can be inconvenient and  cognitively demanding.
 Thus, we associate a TSL with \li{HTML} that provides a simplified notation for writing HTML, shown being used on lines 6-18 of Fig.
@@
 The forward reference will be replaced by the general-purpose term, of type \li{HTML}, generated by the TSL during typechecking.
-Because layout was used as a delimiter, there are no syntactic constraints on the body, unlike with inline forms (Fig.
+Because layout was used as a delimiter, there are no syntactic constraints on the body, unlike with inline literals (Fig.
 \ref{f-delims}).
@@
     val parser = ~
-SGRM      start <- '>body'= attributes start>
+SGRM      start <- '>body'= attributes> start>
         EGRMfn attrs, child => `SQTHTML.BodyElement((%EQTattrsSQT%, %EQTchildSQT%))EQT`
-SGRM      start <- '>style'= attributes EXP>
+SGRM      start <- '>style'= attributes> EXP>
         EGRMfn attrs, e => `SQTHTML.StyleElement((%EQTattrsSQT%, %EQTeSQT%))EQT`
 SGRM      start <- '<'= EXP>EGRM 
-        fn e => `SQT%EQTeSQT% : HTMLEQT`
+        fn e => `SQT%EQTeSQT%EQT`
 \end{lstlisting}
@@
 The layout constraints available are \li{=} (meaning that the leftmost column of the annotated term must be aligned with that of the parent term), \li{>} (the leftmost column must be indented further) and \li{>=} (the leftmost column \emph{may} be indented further).
-We will discuss this formalism further when we formally specify Wyvern's layout-sensitive concrete syntax.
+%We will discuss this formalism further when we formally specify Wyvern's layout-sensitive concrete syntax.
 
@@
 
-We have now seen several examples of TSLs that support splicing.
+%We have now seen several examples of TSLs that support splicing.
 The question then arises: what type should the spliced Wyvern expression be expected to have?
 This is determined by placing the spliced value in a place in the generated AST where its type is known -- on line 11 of Fig.
-\ref{f-htmltype} it is known to be \li{HTML} and on line 13 it is known to be \li{CSS} by the declaration of \li{HTML}, and on line 15, it is known to be \li{HTML} by the use of an explicit ascription.
+\ref{f-htmltype} it is known to be \li{HTML} and on line 13 it is known to be \li{CSS} by the declaration of \li{HTML}.
 When these generated ASTs are recursively typechecked during compilation, any use of a nested TSL at the top-level (e.g.
@@
 \subsection{Formalization}
-A formal and more detailed description can be found in our paper draft.\footnote{\url{https://github.com/wyvernlang/docs/blob/master/ecoop14/ecoop14.pdf?raw=true}} In particular:
+A formal and more detailed description can be found in our paper.\footnote{\url{https://github.com/wyvernlang/docs/blob/master/ecoop14/ecoop14.pdf?raw=true}} In particular:
 \begin{enumerate}
@@
 The key judgements are of the form:
-\[\Gamma'; \Gamma \vdash_\Sigma e \leadsto \hat{e} \Leftarrow \tau~~~~~~~~\text{and}~~~~~~~~\Gamma'; \Gamma \vdash_\Sigma e  \leadsto \hat{e} \Rightarrow \tau\]
+\[\Gamma \vdash_\Theta e \leadsto i \Leftarrow \tau~~~~~~~~\text{and}~~~~~~~~\Gamma \vdash_\Theta e  \leadsto i \Rightarrow \tau\]
 
-Expressions, $e$, elaborate to \emph{literal-free expressions}, $\hat{e}$.
+Expressions, $e$, elaborate to \emph{internal expressions}, $i$ (which do not contain literals).
 This occurs simultaneously with typechecking; the first judgement captures situations where type analysis is occurring, the second type synthesis.
-The typing context, $\Gamma$, is standard, and the named type context, $\Sigma$, associates named types (object types and case types, but not arrow or product types) with their declarations and metadata.
+The typing context, $\Gamma$, is standard, and the named type context, $\Theta$, associates named types with their declarations and metadata.
 
@@
 We show a novel mechanism that ensures hygiene by requiring that the generated AST is closed except for subtrees derived from portions of the user's parse stream that are interpreted as spliced Wyvern expressions or identifiers.
-We formalize this mechanism by splitting the context in our static semantics (in the judgement above, $\Gamma'$ contains variables only allowed in spliced Wyvern expressions).
-We also show how to explicitly refer to local values available in the parser definition (e.g.
+We formalize this mechanism by splitting the context in our static semantics.
+%We also show how to explicitly refer to local values available in the parser definition (e.g.
 helper functions) in a safe way.
@@
 \subsection{Remaining Tasks and Timeline}
-This work has been accepted to ECOOP 2014, having received quite positive reviews.
-Final revisions on the paper text are due on May 12th.% and we will defer these tasks to after the OOPSLA deadline in late March.% If not accepted, we will aim to resubmit to OOPSLA.
-\begin{enumerate}
+This work was published at ECOOP 2014 (and received a Distinguished Paper Award) \cite{TSLs}.
+We have not formally shown how parameterized types work (e.g.
+we can't yet implement list literal syntax uniformly as a library).
+I plan to do this as remaining work in the course of completing the dissertation writing.
+I am currently working with two undergraduates on other extensions to this work, but do not plan to include it in my dissertation.
+% and we will defer these tasks to after the OOPSLA deadline in late March.% If not accepted, we will aim to resubmit to OOPSLA.
+%\begin{enumerate}
 %\item In our current top-level grammar, expressions within parenthesis must still obey layout constraints.
@@
 We will add these simple rules to the grammar.
-\item There are a few presentation issues with our current formalism as written that need to be fixed.
-We must also write down the rules that we omitted for concision in the paper draft in a technical report, as well as provide more complete proofs of the metatheory.
-This might require slightly restricting the recursion in the rule for literals, so that induction can be shown to be well-founded.
-\item Our current static semantics does not support mutually recursive type declarations, but this is necessary for our prelude to typecheck, so we will add support for this.
-In the paper, the formalism is meant to be expository, so we will likely leave this additional complexity for the tech report.
-\item We must further consider aspects of hygiene:
-\begin{enumerate}
-\item We have not yet formalized our mechanism for preventing unintentional variable \emph{shadowing}, only unintentional variable \emph{capture}.
-It should be possible to prevent shadowing by preventing variables from leaking into spliced Wyvern expressions (so that function application is the only way to pass data from a TSL to Wyvern code, as in the \li{Parser} TSL above).
-\item In macro systems like Scheme's, free variables in generated ASTs refer to their bindings within the macro definition, rather than where they are inserted.
+%\item 
+%\item In macro systems like Scheme's, free variables in generated ASTs refer to their bindings within the macro definition, rather than where they are inserted.
 To support this, we have introduced the \li{toast} operator.
@@
 We need to finish this (collaborative work with Darya Kurilova).
-\end{enumerate}
+%\end{enumerate}
 %\item The corpus analysis we conducted was preliminary. We must perform this in a more complete and rigorous manner (collaborative work with Darya Kurilova).
@@
 I don't want to promise that everything will be implemented because that is largely contingent on Benjamin's effort, not mine.}
-\end{enumerate}
+%\end{enumerate}
 %

### 2014-08-27 thesis-proposal/problem.tex

@@
 \section{Motivation}\label{motivation}
-%\begin{quote}\textit{The recent development of programming languages suggests that the simultaneous achievement of simplicity 
-%and generality in language design is a serious unsolved 
-%problem. }
-%- Reynolds, 1970\todo{citations}\end{quote}
+\begin{quote}\textit{The recent development of programming languages suggests that the simultaneous achievement of simplicity 
+and generality in language design is a serious unsolved 
+problem.}
+-- John Reynolds, 1970 \cite{Reynolds70}\end{quote}
 %%One might exp ``generality''
-Specifying and implementing a ``general purpose'' programming language and its associated tools (collectively, a \emph{programming system}) has long been a grand challenge in computing.
-Were a truly general system to emerge, we might expect that its users would rarely  need to develop a new dialect of the system.
-Embeddings of new abstractions using mechanisms built into the language would be suitably reasonable, natural and performant.
-Alas, new language and tool dialects spanning all major language lineages continue to be widely used as vehicles for new abstractions (examples of which we will discuss throughout this work), suggesting that not all useful  abstractions can be seen, when considered comprehensively, as mere modes of use of any contemporary ``general purpose'' programming system.% The problem of achieving both simplicity and generality simultaneously with a single system remains unsolved.
+%Designing a general-purpose programming language along with its supporting tools (collectively, a \emph{programming system}) has long stood as a grand challenge in computing. 
+Achieving both simplicity and generality in a programming language and its associated tools -- collectively, a \emph{programming system} -- requires organizing around a small number of mechanisms that support the realization of many others   as libraries.
+% In such a system, new dialects and revisions of the system would rarely be necessary.
+%Indeed, a preponderence of new dialects is clear evidence against claims of generality.
+%Not all useful abstractions can be  realized, when considered comprehensively, as libraries.% The problem of achieving both simplicity and generality simultaneously with a single system remains unsolved.
+Much progress has been made in the decades since Reynolds' statement above, but a programming system that comprehensively achieves these goals remains elusive, as evidenced by the fact that across all language lineages,  new language and tool {dialects} remain common vehicles for new features and abstractions, both in research and practice.
 
-We observe that new dialects are often motivated by the desire to introduce new syntax, strengthen the language's static semantics or provide specialized editor services, all aspects of existing systems that appear fixed when considered from within (that is, from the perspective of a library provider).
+We observe that these new dialects are often motivated by the desire for new concrete syntax, new type system fragments, or new editor services, all {features} of programming systems that library providers usually do not have substantial control over.
 %, encouraged  historically  by the availability of tools like compiler generators and,  more recently, language workbenches \cite{workbenches} and DSL frameworks \cite{dsl}.
 Unfortunately, taking this approach makes it substantially more difficult for clients to import high-level abstractions orthogonally.
-It follows then that {language-integrated mechanisms} that allow library providers to extend these features of the system could strengthen its claim to generality by decreasing the need for new language and tool dialects.
-However, such mechanisms must be introduced with care, because decentralizing control over such fundamental aspects of the system could weaken its metatheory and permit ambiguities and conflicts when two libraries are combined (collectively, \emph{safety issues}).
-It could also make it more difficult to define new functions over expression forms in the language (the \emph{expression problem}) and make it more difficult to  understand the meaning of programs.
-In this thesis, we aim to show that by associating system extensions with types (forming what we call \emph{active types}) rather than new expression forms, and using a \emph{bidirectional type system} that enforces critical  abstraction barriers between extensions, we can  give abstraction providers the ability to orthogonally implement new syntax, type systems and editor services from within libraries, rather than as new language and tool dialects, while avoiding safety issues and retaining  an essentially conventional typing discipline.
+Introducing mechanisms that decentralize control over these features could increase generality, but they must constructed with care, because giving away too much control could weaken the metatheory of the system.
+Moreover, they must come equipped with \emph{modular reasoning principles} for it to be practical for library clients to rely upon them.
+Having to manually resolve conflicts between libraries or reestablish important properties of a program any time a new library is imported introduces undesirable complexity.
+%It can also make it more difficult to define new sorts of tools for working with the language because they cannot rely on there being a fixed abstract syntax (the canonical \emph{expression problem}).
+For programmers,
+%It can make it difficult to understand and reason about the type of an unfamiliar term, a critical facet of program comprehension (the \emph{typing discipline problem}).
+
+In this thesis, we aim to design mechanisms that handle these issues by embracing the already widely accepted notion that \emph{type constructors} (or more concisely, \emph{tycons}) can serve as the fundamental organizational unit of a programming system.
+We show how to delegate control over certain aspects of the concrete syntax, external type system and editor services to user-defined functions associated with type constructors, together forming what we call \emph{active type constructors}.
+This permits library-based realizations of features that today would need to be built into a system by its central designers.
+By layering these mechanisms over a fixed typed internal language and constraining these functions appropriately, we retain key  metatheoretic properties and arrive at powerful modular reasoning principles.
+% while maintaining key metatheoretic properties and modular reasoning principles.
+% By keeping the abstract syntax fixed and layering these mechanisms atop a fixed typed internal language (essentially lifting the first stages of compilation into the language), we avoid most aspects of the expression problem.
+%By delegating control to user-defined functions provided alongside user-defined type constructors (forming \emph{active type constructors}), we can give library providers substantially more control over these features of the system.
+%We structure each mechanism that we introduce as a \emph{bidirectional type system} combined with an  \emph{elaboration semantics} targeting a fixed typed internal language. 
+%By constraining these functions and enforcing critical abstraction barriers between extensions, we will ensure key metatheoretic properties of the language and system components and guarantee that extensions can be reasoned about modularly.
 %We call user-defined types that introduce new features into the system in this way \emph{active types}.
@@
 \subsection{Motivating Example: Regular Expressions}\label{regex}
-To make the problems we aim to address more concrete, we begin with a simple example that we will return to throughout this work.
-\emph{Regular expressions} are commonly used to express patterns in strings (e.g.
-DNA sequences) \cite{Thompson:1968:PTR:363347.363387}.
-Programmers who work with regular expressions regularly might want a programming system supporting features like:
+To make the use cases we seek to address more concrete, let us begin with a simple example that we will return to throughout this proposal.
+\emph{Regular expressions} are commonly used to capture patterns in strings (e.g.
+motifs in DNA sequences) \cite{Thompson:1968:PTR:363347.363387}.
+Programmers who work with regular expressions would benefit from features like these:
 
 \begin{enumerate}
-\item \textbf{Syntax for pattern literals}.
-An ideal syntax would permit us to express patterns in a concise, conventional manner.
-For example, The BisI restriction enzyme cuts DNA whenever it sees the pattern \texttt{GC\textit{N}GC}, where \texttt{\textit{N}} represents any base.
-We would want to express it, perhaps, as follows (using curly braces to splice one pattern into another):
+\item \textbf{Concrete syntax for pattern literals}.
+An ideal syntax would permit programmers to express patterns in the concise, conventional manner.
+For example, consider a bioinformatics  application: the BisI restriction enzyme cuts DNA when it sees the motif \texttt{GC\textit{N}GC}, where \texttt{\textit{N}} is any base.
+We would want to express such patterns with syntax like this, using curly braces to splice one pattern into another and parentheses to delimit captured groups:
 \begin{lstlisting}[numbers=none]
 let N : Pattern = <SURLA|T|G|CEURL>
-let BisI : Pattern = <SURLGC{EURLNSURL}GCEURL>\end{lstlisting}
+let BisI : Pattern = <SURLGC({EURLNSURL})GCEURL>\end{lstlisting}
 %The cognitive load of reading and writing patterns is low and patterns are parsed once at compile-time. 
-We would want malformed patterns to result in intelligible {compile-time} errors.
-\item A \textbf{type system} that ensures that key invariants related to regular expressions are statically maintained:
+Malformed patterns would result in intelligible {parser} errors.
+\item A \textbf{type system} that ensures that key invariants related to regular expressions are statically maintained. It might statically ensure that: 
 	\begin{enumerate}
-	\item only other patterns and properly escaped strings are spliced into a pattern, to avoid splicing errors and injection attacks \cite{owasp2013, Bravenboer:2007:PIA:1289971.1289975}
-	\item out-of-bounds backreferences to a captured group are not used \cite{spishak2012type}
+	\item only patterns, or properly escaped strings, are spliced into a pattern literal, to avoid splicing errors and injection attacks \cite{owasp2013, Bravenboer:2007:PIA:1289971.1289975}
+	\item out-of-bounds backreferences to a captured group do not occur \cite{spishak2012type}
 	\item string processing operations do not lead to a string that is malformed, when well-formedness can be captured as membership in a regular language \cite{fulton-thesis,HosoyaVouillonPierce2000ICFP}
@@
 %By performing this step ahead-of-time, these dangers can be avoided.
-\item \textbf{Editor services} to support syntax highlighting, documentation, interactive testing and pattern extraction from example strings.
+\item \textbf{Editor services} to support syntax highlighting, consulting relevant documentation, interactive testing and pattern extraction from example strings.
 \end{enumerate}
@@
 %In a conventional \emph{monolithic} programming system, support for each of these features would need to be built into the language and tools. 
-No system today builds in support for all of the features enumerated above in their strongest form, so library providers must provide support for regular expressions by leveraging  general-purpose mechanisms.
-Unfortunately, it is impossible to completely define the syntax and the specialized static semantics referenced above in terms of general-purpose notations and abstractions, so library providers need to compromise.
+No system today provides built-in support for all of the features enumerated above in their strongest form, so library providers must leverage  general-purpose mechanisms.
+%Unfortunately, it is impossible to completely define the syntax and the semantics referenced above using contemporary mechanisms, so library providers need to compromise.
 
-The most common strategy is to ask clients to enter patterns as strings, deferring their parsing, typechecking, compilation and use all to run-time.
-This provides only a partial approximation to feature 1 (due to clashes between string escape sequences and regular expression syntax).
-None of the static guarantees can be provided in this way, leading to run-time exceptions (even in well-tested code \cite{spishak2012type}), and logic errors that cannot even be caught at run-time, some of which can lead to security vulnerabilities (due to injection attacks \cite{owasp2013}).
+The most common strategy is to ask clients to introduce patterns as strings, deferring their  parsing, compilation and use to run-time.
+This provides only a partial approximation to feature 1 due to syntactic clashes between string escape sequences and regular expressions.
+Parsing errors due to these clashes have been observed to be difficult for even experienced programmers to diagnose \cite{Omar:2012:ACC:2337223.2337324}.
+Moreover, none of the static guarantees can be provided, leading to run-time exceptions (common even in well-tested code \cite{spishak2012type}), and mistakes that are not caught even at run-time, some of which can lead to security vulnerabilities due to injection attacks \cite{owasp2013}.
 It also introduces performance overhead due to run-time parsing and compilation of patterns, and redundant run-time checks.
+Staging these computations can shift some of the errors and cost to compile-time, albeit with some additional semantic and syntactic complexity, but the issues related to the fact that a regular expression is not a string (e.g.
+syntactic clashes and injection attacks) persist.% parsing can move some errors to run-time, albeit at additional syntactic cost, but it also does not straightforwardly handle splicing-related issues, as we will discuss later.
 
-Requiring that clients instead introduce patterns directly using general-purpose constructs like sums or products (as exposed by, e.g., functional datatypes or objects) or operations over an abstract type, rather than via strings, provides only feature 2a at the expense of the weak approximation to feature 1 that the use of string literals provided.
+A more semantically justifiable approach is to dispense with string representations entirely and directly work with an encoding of regular expression patterns using  general-purpose constructs like sums and  products (e.g.
+as exposed by functional datatypes or objects), or using an abstract type.
+This provides feature 2a while sacrificing feature 1 (such an encoding affords only an approximation of the verbose abstract syntax of patterns, not its concrete syntax).
+Advanced invariants like those described in features 2b and 2c are not directly tracked by such encodings and require solutions beyond those commonly found in modern simply-typed languages (we will return to existing approaches  in Sec.
+\ref{att}).%Explicitly constructing a regular expression by applying datatype constructors is unacceptably verbose.
 
-None of these approaches address the issue of tool support (feature 3).
-Regular expressions are not trivial to work with, and we have found that tool support can be quite helpful, even for professional programmers \cite{Omar:2012:ACC:2337223.2337324}.
-Indeed, a number of tools are available online that provide services like syntax highlighting, testing and explanations of patterns \cite{regexr} and pattern extraction from examples (e.g.
+None of these approaches address the important issue of tooling (feature 3).
+A number of tools are available online that provide services ranging from simple syntax highlighting to interactive testing and explanations of complex patterns \cite{regexr} and pattern extraction from examples (e.g.
 \cite{_txt2re:_????}).
-Unfortunately, tools that must be accessed externally are difficult to discover and are thus used infrequently \cite{Murphy-Hill:2011:PIE:1958824.1958888,Campbell:2008:DRT:1636642.1636651,Omar:2012:ACC:2337223.2337324}.
-We found that they are also less usable than editor-integrated tools %(e.g.
+We have found that programmers benefit substantially from their use \cite{Omar:2012:ACC:2337223.2337324}.
+Unfortunately, evidence suggests that tools that must be accessed externally are difficult to discover and are thus used infrequently \cite{Murphy-Hill:2011:PIE:1958824.1958888,Campbell:2008:DRT:1636642.1636651,Omar:2012:ACC:2337223.2337324}.
+We have found that they are also less usable than editor-integrated tools %(e.g.
 \cite{IntelliJRegexp})
 because they require the programmer to switch contexts and cannot make use of the code context \cite{Omar:2012:ACC:2337223.2337324}.
-Working with high-level abstractions in a ``low-level'' editor is thus awkward and error-prone, just like attempting to work with high-level abstractions in a low-level language.
-For these reasons,   we consider editor behavior as within the scope of an abstraction's specification.
+Working with high-level abstractions like these in a ``low-level'' editor is thus awkward and error-prone, just like attempting to work with high-level abstractions directly in terms of an encoding.
+For these reasons,   we will consider editor services as within the scope of an abstraction specification.
 \subsection{Language-External Approaches}\label{external-approaches}
@@
 \vspace{-20px}
-\caption{\small (a) When taking a language-external approach, new features are packaged together into separate languages and tools, causing problems with orthogonality and client compatibility (described in the text).
-(b) When taking a language-integrated approach, there is one extensible host language and the compile-time and edit-time logic governing new constructs is expressed within ``active libraries''.
-If the extension mechanism precludes conflicts by construction, the problems of orthogonality and client compatibility are avoided.}
+\caption{\small (a) When taking a language-external approach, new features are packaged together into separate languages and tools.
+(b) When taking a language-integrated approach, there is one extensible host language and the compile-time and edit-time logic governing new constructs is expressed within ``active libraries''.}
 \label{approaches}
@@
 
-When the syntax and semantics of a system must be extended to fully realize a new feature providers typically take a \emph{language-external approach}, either by developing a new system dialect (supported by \emph{compiler generators} \cite{brooker1963compiler}, \emph{language workbenches} \cite{erdweg2013state}, \emph{DSL frameworks} \cite{fowler2010domain}, or simply be forking an existing codebase), or by using an extension mechanism for a {particular} compiler\footnote{Compilers that modify, or allow modification of, the semantics of their base language, rather than simply permitting semantics-preserving optimizations, should be considered a pernicious means for creating new languages.
+In situations, like these, where a library-based approach is not satisfying, providers must take a \emph{language-external approach}, either by developing a new system dialect (supported by \emph{compiler generators} \cite{brooker1963compiler}, \emph{language workbenches} \cite{erdweg2013state}, \emph{DSL frameworks} \cite{fowler2010domain}, or simply by forking an existing codebase), or by using an extension mechanism for a {particular} compiler\footnote{Compilers that modify, or allow modification of, the semantics of their base language, rather than simply permitting semantics-preserving optimizations, should be considered a pernicious means for creating new language dialects.
 That is, some programs that purport to be written in C, Haskell or Standard ML are actually written in compiler-specific dialects of these languages.}, editor or other tool.
-For example, a researcher interested in providing regular expression related features (let us refer to these collectively as \texttt{R}) might design a new system with built-in support for these, perhaps basing it on an existing system containing some general-purpose features (\texttt{G}).
+For example, a researcher interested in providing the regular expression related features just described (let us refer to these collectively as \texttt{R}) might design a new system with built-in support for them, perhaps basing it on an existing system containing some general-purpose features (\texttt{G}).
 A different researcher developing a new language-integrated parallel programming abstraction (\texttt{P}) might  take the same approach.
@@
 This results in a collection of distinct systems, as diagrammed in Figure \ref{approaches}a.
-This might be entirely sufficient if developing a proof of concept is the only goal, but it makes it difficult to achieve adoption of these abstractions (even by other researchers) related to \textbf{orthogonality} and \textbf{client compatibility}.% This has limited the broad adoption of these kinds of innovations.%This latter method couples the semantics of the feature to the implementation details of a particular tool.
+Although often justified as ``proofs of concept'', this is a rather weak notion because no abstraction will be used entirely in isolation and such a construction gives us no rigorously justifiable  reason to believe that it can safely and naturally coexist with others specified or implemented in the same way.
+That is, one must either use the system containing features \texttt{G+R}, \texttt{G+P} or \texttt{G+Q}.
+There is no system containing \texttt{G}, \texttt{R}, \texttt{P} and \texttt{Q} in other combinations, and merging the systems containing each separately can be non-trivial because there can be serious interference and safety issues, as we will discuss at length throughout this thesis.
+% More specifically, there is no guarantee of \textbf{orthogonality} or even \textbf{interoperability}.% This has limited the broad adoption of these kinds of innovations.%This latter method couples the semantics of the feature to the implementation details of a particular tool.
 Because the use of one implementation entails a different semantics for the feature than another, the extended tool acts, \emph{de facto}, as a distinct system for our purposes.
 
-\paragraph{Orthogonality} Features implemented by language-external means cannot be adopted individually, but instead are only available coupled to a fixed collection of other features.
-This makes adoption more costly when these incidental features are  not desirable or insufficiently developed, or when the features bundled with a different language or tool are simultaneously desirable.
-That is, one must either use the system containing features \texttt{G+R}, \texttt{G+P} or \texttt{G+Q}.
-There is no system containing \texttt{G}, \texttt{R}, \texttt{P} and \texttt{Q} in other combinations, and merging the systems containing each separately can be non-trivial because, even in cases where a common mechanism has been used, there are serious interference and safety issues, as we will discuss at length throughout this thesis.
+%\paragraph{Orthogonality} Features implemented by language-external means like these cannot be adopted individually, but instead are only available coupled to a fixed collection of other features.
+This makes adoption more costly when these incidental features are  not desirable or are insufficiently developed (``toy languages''), or when the features bundled with a different language or tool are simultaneously desirable.
 
-Recent evidence indicates that this is one of the major barriers preventing research from being driven into practice.
-For example, developers prefer language-integrated parallel programming abstractions with stronger semantic guarantees, more efficient implementation and more natural syntax to library-based approximations when all else is equal \cite{cave2010comparing}, but library-based approximations are more widely adopted because  ``parallel programming languages'' privilege only a few chosen  abstractions at the language level.
-This is problematic, because different abstractions are seen as more appropriate in different situations \cite{Tasharofi:2013rc}.
-Moreover,  parallel programming is rarely the only concern relevant to clients outside of a classroom setting.
-Support for regular expressions, for example, would be simultaneously desirable for processing large amounts of genomic data in parallel, but using these features together in the same compilation unit would be difficult or impossible.
-Indeed, switching to a ``parallel programming language'' would likely make it \emph{more} difficult to use regular expressions, as these are likely to be less well-developed in a specialized language than in an established general-purpose language.
-The intuition was perhaps most succinctly expressed by a participant in a recent study by Basili et al.
+Recent evidence indicates that this is one of the major barriers preventing research on programming languages from being driven into practice.
+For example, developers prefer high-level language-integrated parallel programming abstractions that provide  strong semantic guarantees when all else is equal \cite{cave2010comparing}, but library-based approximations are far more widely adopted because  ``parallel programming languages'' privilege only a few chosen  abstractions at the language level.
+This is problematic because different abstractions are seen as more appropriate in different situations \cite{Tasharofi:2013rc}.
+Moreover,  parallel programming is rarely the only concern relevant to clients outside of a classroom or research setting.
+Support for regular expressions, for example, would be simultaneously desirable for processing large amounts of genomic data in parallel, but using these features together in the same compilation unit would be difficult or impossible if implemented using language-external means.
+Indeed, switching to a ``parallel programming language'' would likely make it \emph{more} difficult to use regular expressions, as these are likely to be less well-developed in a specialized language than in an established general-purpose language.% This intuition was perhaps most succinctly expressed by a participant in a recent study by Basili et al.
 \cite{basili2008understanding}:  ``I hate MPI, I hate C++.
@@
 
-\paragraph{Client Compatibility} Even in cases where, for each component of a software system, there was a programming system considered entirely satisfactory by its developers, there would remain a problem at the interface between its components.
-An interface that exposes externally a specialized construct particular to one language (e.g.
+\paragraph{Interoperability} Even in cases where, for each component of a software system, a programming system considered entirely satisfactory by its developers is available (e.g.
+a team goes through the trouble of implementing \verb|G+R+P| by reading papers about \verb|G+R| and \verb|G+P| and disentangling orthogonality-related issues), there remains a problem at any interface between  components written using a different combination of features.
+An interface that  externally exposes a specialized construct particular to one language (e.g.
 a function that requires a quantity having a particular unit of measure) cannot necessarily be safely and naturally consumed from another language (e.g.
@@
 Tool support is also lost when calling into different languages.
-We call this the \emph{client compatibility problem}: code written by clients of a certain collection of features cannot always interface with code written by clients of a different collection  in a safe, performant and natural manner.
+%We call this the \emph{interoperability problem}.
+% programs written by clients of a certain collection of features cannot always interface with programs written by clients of other features  in a safe, performant and natural manner.
 
-One strategy taken by proponents of the {language-oriented approach} to abstraction development \cite{journals/stp/Ward94} to partially address the client compatibility problem is to  target an established intermediate language and use its constructs as a common language for communication between components written in different languages.
+One strategy taken by proponents of a {language-oriented approach} \cite{journals/stp/Ward94} to partially address the interoperability problem is to  target an established intermediate language and use its constructs as a common language for communication between components written in different languages.
 Scala \cite{200464/IC} and F\# \cite{pickering2007foundations} are examples of prominent general-purpose languages that have taken this approach, and most DSL frameworks also rely on this strategy.
-As indicated in Figure \ref{approaches}a, this only enables client compatibility in one direction.
-Calling into the common language becomes straightforward and safe, but calling in the other direction, or between the languages sharing the common target, does not, unless these languages are only trivially different from the intermediate language.
+As indicated in Figure \ref{approaches}a, this only enables interoperability in one direction.
+Calling into the common language becomes straightforward and safe, but calling in the other direction, or between the languages sharing the common target, does not, unless these languages are only trivially different from the common language.
 
@@
 The F\# type system also includes support for checking that units of measure are used correctly \cite{syme2012expert, kennedy1994dimension}, but this more specialized static invariant is left entirely unchecked at language boundaries.
-Exposing functions that operate over datatypes, tuples and values having units of measure is not recommended when a component ``might be used'' from another language \cite{syme2012expert} because it is awkward to construct and consume these from other languages without the convenient primitive operations (e.g.
+Indeed, guidelines for F\# suggest that exposing functions that operate over values having units of measure, datatypes or tuples is not recommended when a component ``might be used'' from another language \cite{syme2012expert} because it is awkward to construct and consume these from other languages without the convenient primitive operations (e.g.
 pattern matching) and syntax that F\# includes.
 SML.NET prohibits exposing such types at component boundaries altogether.
-It cannot naturally consume F\# data structures, despite having a rather similar syntax and semantics in most ways (both languages directly descend from ML).
+Moreover, it also cannot naturally consume F\# data structures, despite having a rather similar syntax and semantics in most ways (both languages directly descend from ML).
 %In Scala, traits that have default method implementations are difficult to implement from Java or other JVM languages and the workaround can break if the trait is modified \cite{scalatraitinterop}. 
@@
 \subsection{Language-Integrated Approaches}\label{language-integrated-approaches}
-We argue that, due to these problems with orthogonality and client compatibility, taking a language-external approach to realizing a new feature should be considered harmful and avoided whenever possible.
-The goal of the research being proposed here is to design \emph{language-integrated extension mechanisms} that give providers the ability to define, within libraries, new features that have previously required central planning,
+We argue that, due to these problems with orthogonality and interoperability, taking a language-external approach to realizing a new feature should be considered harmful and avoided whenever possible.
+The goal of the research being proposed here is to design \emph{language-integrated extension mechanisms} that give providers the ability to define, within libraries, new features that, like those above, have previously required central planning,
 %\footnote{One might compare today's programming systems to  {centrally-planned} economies, whereas extensible\- systems more closely resemble modern market economies.
@@
 We leave further development of this analogy to the reader.}
-so that language-external approaches are less frequently necessary.
-More specifically, we will show how control over aspects of the \textbf{syntax}, \textbf{type system} and \textbf{editor services} can be delegated to user-defined logic distributed in {libraries}, as illustrated in Figure \ref{approaches}b.
-Such libraries have been called \emph{active libraries}  \cite{activelibraries} because, rather than being passive clients of features already available in the system, they contain logic invoked by the components of the programming system while a program is being edited or compiled to provide new features.
-Features implemented within active libraries can be imported as needed, unlike features implemented by external means, seemingly avoiding the problems of orthogonality and client compatibility.
+so that language-external approaches are less frequently necessary, as illustrated in Figure \ref{approaches}b. 
+Such libraries have been called \emph{active libraries}  \cite{activelibraries} in that they are not passive clients of features already available in the system.
+Features implemented within active libraries can be imported individually, unlike features implemented by external means, giving us a potential means to avoid the problems of orthogonality and interoperability.
 
-We must proceed with caution, however: critical issues having to do with {safety} must be overcome before language-integrated extension mechanisms can be integrated into a system.
-If too much control over  these core aspects of the system is given  to developers, the system may become quite unreliable.
+We must proceed with caution, however: critical issues having to do with {safety} must be overcome before language-integrated extension mechanisms can be introduced into a system.
+If too much control over  these core features of the system is given  to developers, the system may become quite unreliable.
 %For example, an extension could weaken important metatheoretic guarantees previously provided by the system. 
 Type safety, for example, may not hold if the static and dynamic semantics of the language can be modified or extended arbitrarily from within libraries.
-Furthermore, even if extensions can be shown not to cause such problems in isolation, there may still be conflicts between extensions that could weaken their semantics, leading to subtle problems that only appear when two extensions are used together.
-As a simple example, if two active libraries introduce the same syntactic form but back it with differing (but individually valid) semantics, the issue would only manifest itself when both libraries were imported within the same scope.
-Resolving these kinds of ambiguities requires significantly more expertise with parser technology than using the syntax itself does.
-These (and other more serious) safety and ambiguity issues have plagued previous attempts to design language-integrated extensibility mechanisms.
-We will briefly review some of these attempts below.% To prevent them, our mechanisms will organize  extension logic around types to guarantee that extensions are both safe in isolation and also safely composable in any combination.
+Furthermore, even if extensions can be shown not to cause such problems in isolation, there may still be conflicts between extensions that could weaken their semantics, leading to subtle problems that only appear when two extensions are used together, thwarting any attempts to reason modularly about particular programs and the system as a whole.
+%As a simple example, if two active libraries introduce the same syntactic form but back it with differing (but individually valid) semantics, this ambiguity  would only manifest itself when both libraries were imported within the same scope.
+%Resolving these kinds of ambiguities requires significantly more expertise with parser technology than using the syntax itself does.
+These issues have plagued previous attempts to design language-integrated extensibility mechanisms.
+We will briefly review some of these attempts below, then return to our approach.% To prevent them, our mechanisms will organize  extension logic around types to guarantee that extensions are both safe in isolation and also safely composable in any combination.
 

### 2014-08-27 thesis-proposal/timeline.tex

@@
 \end{quote}
-But we take issue with this unbounded \emph{principle of tolerance}, which he explained more directly as follows \cite{CarnapR:logsl}:
+He explained his \emph{principle of tolerance} further as follows \cite{CarnapR:logsl}:
 \begin{quote}
@@
 But he was also a prominent member of the Vienna circle, which sought a ``unified science'' the purpose of which was "to link and harmonise the achievements of individual investigators in their various fields of science" by the construction of a "constitutive system" \cite{hahn1929scientific}.
-These two programs were in conflict: were the Vienna circle's ``constitutive system'' to become na\"ively tolerant of the inference rules (what Carnap called L-rules) developed by individual investigators, it would find that this would then deny them autonomy of reasoning.
-According to social scientist Raz, the primary justification of tolerance is to maximize
-individual autonomy in a multi-party context \cite{raz1988autonomy}.
+These two programs were in conflict: were the Vienna circle's ``constitutive system'' to become na\"ively tolerant of the inference rules (what Carnap called L-rules) developed by individual investigators, it would find that this would then deny them autonomy of reasoning.% According to social scientist Raz, the primary justification of tolerance is to maximize
+%individual autonomy in a multi-party context \cite{raz1988autonomy}.
 Yossie Nehushtan argues that a liberal  society ``should not tolerate anything that denies the justifications of tolerance'' \cite{nehushtan2007limits}.
-Thus, we propose a more cooperative \emph{principle of reciprocal tolerance}:
+Thus, we propose a \emph{principle of reciprocal tolerance}:
 \begin{quote}
@@
 Everyone is at liberty to build their own logical fragment, i.e.
-their own language extensions, as long as these can be implemented within a framework that ensures, by formal means instead of philosophical argument, that autonomously derived reasoning principles are conserved when fragments are combined, in \emph{any} combination.
+their own language extensions, as long as these can be implemented within a framework that ensures, by formal means instead of philosophical argument, that autonomously derived reasoning principles are conserved when fragments are combined.
 \end{quote}
@@
 
-\section{Timeline}
-To summarize, we plan on completing the work in this thesis per the following schedule:
-\begin{itemize}
-\item The work on type-specific languages has been accepted to ECOOP 2014.
+% \section{Summary of }
+% To summarize, we plan on completing the work in this thesis per the following schedule:
+% \begin{itemize}
+% \item The work on type-specific languages has been accepted to ECOOP 2014.
 The deadline for final revisions is May 12, so we will complete the remaining work, including writing a technical report, by then.
 This should be sufficient for the chapter on Wyvern in the dissertation.
-\item The work on @$\lambda$ and active embeddings is conceptually complete.
+% \item The work on @$\lambda$ and active embeddings is conceptually complete.
 We plan to complete the technical details and exposition and submit it to POPL in July.
 This should be sufficient for the chapter on @$\lambda$ in the dissertation.
-\item The work described the core of Ace was submitted to OOPSLA 2014 on Mar.
+% \item The work described the core of Ace was submitted to OOPSLA 2014 on Mar.
 25.
@@
 Together, these will form the chapters on Ace in the dissertation.
-\item The work on active code completion has been completed and published at ICSE 2012 \cite{Omar:2012:ACC:2337223.2337324}.
+% \item The work on active code completion has been completed and published at ICSE 2012 \cite{Omar:2012:ACC:2337223.2337324}.
 The remaining work is expository and will be completed in the course of writing the Graphite chapter of the dissertation.
-\end{itemize}
+% \end{itemize}

### 2014-09-01 thesis-proposal/problem.tex
small revisions to proposal

@@
 Much progress has been made in the decades since Reynolds' statement above, but a programming system that comprehensively achieves these goals remains elusive, as evidenced by the fact that across all language lineages,  new language and tool {dialects} remain common vehicles for new features and abstractions, both in research and practice.
+%Library implementations are not yet satisfying.
 
-We observe that these new dialects are often motivated by the desire for new concrete syntax, new type system fragments, or new editor services, all {features} of programming systems that library providers usually do not have substantial control over.
+We observe that new system dialects are often motivated by a desire for new concrete syntax, new type system fragments, or new editor services, all {features} of programming systems that library providers usually do not have substantial control over.
 %, encouraged  historically  by the availability of tools like compiler generators and,  more recently, language workbenches \cite{workbenches} and DSL frameworks \cite{dsl}.
 Unfortunately, taking this approach makes it substantially more difficult for clients to import high-level abstractions orthogonally.
-Introducing mechanisms that decentralize control over these features could increase generality, but they must constructed with care, because giving away too much control could weaken the metatheory of the system.
-Moreover, they must come equipped with \emph{modular reasoning principles} for it to be practical for library clients to rely upon them.
-Having to manually resolve conflicts between libraries or reestablish important properties of a program any time a new library is imported introduces undesirable complexity.
+Introducing mechanisms that decentralize control over these features could decrease the need for new dialects and thus increase generality.
+But this must done with care, because giving away too much control could weaken important metatheoretic guarantees.
+Moreover, such  mechanisms must come equipped with \emph{modular reasoning principles} for it to be practical for library clients to rely upon them.
+Having to manually resolve conflicts between libraries or reestablish important properties of a program whenever a new library is imported introduces undesirable complexity.
 %It can also make it more difficult to define new sorts of tools for working with the language because they cannot rely on there being a fixed abstract syntax (the canonical \emph{expression problem}).
@@
 The most common strategy is to ask clients to introduce patterns as strings, deferring their  parsing, compilation and use to run-time.
-This provides only a partial approximation to feature 1 due to syntactic clashes between string escape sequences and regular expressions.
-Parsing errors due to these clashes have been observed to be difficult for even experienced programmers to diagnose \cite{Omar:2012:ACC:2337223.2337324}.
+This provides only a partial approximation to feature 1 due to clashes between string escape sequences and regular expressions (parsing errors due to these clashes have been observed to be difficult for even experienced programmers to diagnose \cite{Omar:2012:ACC:2337223.2337324}).
 Moreover, none of the static guarantees can be provided, leading to run-time exceptions (common even in well-tested code \cite{spishak2012type}), and mistakes that are not caught even at run-time, some of which can lead to security vulnerabilities due to injection attacks \cite{owasp2013}.
@@
 as exposed by functional datatypes or objects), or using an abstract type.
-This provides feature 2a while sacrificing feature 1 (such an encoding affords only an approximation of the verbose abstract syntax of patterns, not its concrete syntax).
-Advanced invariants like those described in features 2b and 2c are not directly tracked by such encodings and require solutions beyond those commonly found in modern simply-typed languages (we will return to existing approaches  in Sec.
+This provides feature 2a while sacrificing feature 1 (such an encoding affords only an approximation of the verbose abstract syntax of patterns, not their concrete syntax).
+Advanced invariants like those described in features 2b and 2c are not directly tracked by such encodings and require solutions beyond those commonly found in modern simply-typed languages (we will return to this  in Sec.
 \ref{att}).%Explicitly constructing a regular expression by applying datatype constructors is unacceptably verbose.
 
-None of these approaches address the important issue of tooling (feature 3).
+None of these approaches address the important issue of tool support (feature 3).
 A number of tools are available online that provide services ranging from simple syntax highlighting to interactive testing and explanations of complex patterns \cite{regexr} and pattern extraction from examples (e.g.
@@
 Working with high-level abstractions like these in a ``low-level'' editor is thus awkward and error-prone, just like attempting to work with high-level abstractions directly in terms of an encoding.
-For these reasons,   we will consider editor services as within the scope of an abstraction specification.
+For these reasons,   we see editor services as within the scope of an abstraction specification.
 \subsection{Language-External Approaches}\label{external-approaches}

### 2014-09-07 att-esop15/att-esop15.tex
Added initial commit of ESOP15 and updated abstract

--- 2014-07-09 att-popl15/att-popl15.tex
@@
-\documentclass[9pt,preprint]{sigplanconf}
-\usepackage{amsmath,tabu}
-\usepackage{cancel}
+\documentclass{llncs}
+\usepackage{amsmath}
+\usepackage{llncsdoc}
 \usepackage{amssymb} 
-\usepackage{amsthm}
-\usepackage{graphicx}
+%\usepackage{amsthm}
+\usepackage{ stmaryrd }
+\usepackage{mathpartir}
 \usepackage[usenames,dvipsnames,svgnames,table]{xcolor}
 \newcommand{\copyleft}{\reflectbox{\sffamily\copyright}}
-\usepackage{ stmaryrd }
-\usepackage{mathpartir}
+%\input{macros-atlam}
+\input{macros-catlam}
 %\usepackage{cite}
 %\renewcommand{\citepunct}{,\,} % IEEEtran wants to use ],\,[ for this but that looks dumb...
-\input{macros-atlam}
-\input{macros-catlam}
-%\usepackage{sfmath}
+
 \renewcommand{\ttdefault}{txtt}
@@
 
-\addtolength{\floatsep}{-8px}
-
-\newcommand{\compresslist}{
-  \vspace{-1em}
-  \setlength{\itemsep}{1pt}
-  \setlength{\parskip}{0pt}
-  \setlength{\parsep}{0pt}
-}
-
 \usepackage{url}
@@
 \usepackage{todonotes}
-
-\lefthyphenmin=10
+\lefthyphenmin=7
 \sloppy
 
-\newcommand{\F}[1]{{\sf #1}~}
-\newcommand{\FF}[1]{{\sf #1}}
-%\usepackage{mathptmx}
-%\usepackage[T1]{fontenc}
-
-\usepackage{microtype}
-    \renewcommand{\topfraction}{0.9}  % max fraction of floats at top
-    \renewcommand{\bottomfraction}{0.9} % max fraction of floats at bottom
-\newcommand{\rulename}[1]{({\textsc{\textsl{#1}}})}
-\def \TirNameStyle #1{\small\rulename{#1}}
-\renewcommand{\MathparLineskip}{\lineskiplimit=.6\baselineskip\lineskip=.6\baselineskip plus .2\baselineskip}
-
-\newcommand{\keyrule}{{\color{red}~$\star$}}
-\newcommand{\minput}{}
-\newcommand{\moutput}{^{\color{gray}{+}}}
-
-\newtheorem{theorem}{Theorem}
-\newtheorem{lemma}{Lemma}
-\newtheorem{prop}{Proposition}
-
 \begin{document}
-
-\conferenceinfo{-}{-} 
-\copyrightyear{-} 
-\copyrightdata{[to be supplied]}
-
-%\titlebanner{{\tt \textcolor{Red}{{\small Under Review -- distribute within CMU only.}}}}        % These are ignored unless
-%\preprintfooter{Distribute within CMU only.}   % 'preprint' option specified.
-
+\mainmatter  % start of an individual contribution
+
+% first the title is needed
 \title{Modular Type Constructors}
-\subtitle{Type System Fragments as Safely Composable Libraries}
-%\subtitle{Conservatively Extending a Type System From Within}
-\authorinfo{}{}{}
-%\authorinfo{Cyrus Omar\and Nathan Fulton\and Jonathan Aldrich}
- %         {Carnegie Mellon University}
-  %        {\{comar, nfulton, aldrich\}@cs.cmu.edu}
-
+\subtitle{Conservatively Composing Typed Language Fragments}
+% a short form should be given in case it is too long for the running head
+\titlerunning{Lecture Notes in Computer Science: Authors' Instructions}
+
+% the name(s) of the author(s) follow(s) next
+%
+% NB: Chinese authors should write their first names(s) in front of
+% their surnames. This ensures that the names appear correctly in
+% the running heads and the author index.
+%
+\author{Cyrus Omar%
+\and Jonathan Aldrich}
+%
+\authorrunning{Lecture Notes in Computer Science: Authors' Instructions}
+% (feature abused for this document to repeat the title also on left hand pages)
+
+% the affiliations are given next; don't give your e-mail address
+% unless you accept that it will be published
+\institute{Carnegie Mellon University, Pittsburgh, PA 15213, USA\\
+\texttt{\{comar,aldrich\}@cs.cmu.edu}
+}
+
+%
+% NB: a more complex sample for affiliations and the mapping to the
+% corresponding authors can be found in the file "llncs.dem"
+% (search for the string "\mainmatter" where a contribution starts).
+% "llncs.dem" accompanies the document class "llncs.cls".
+%
+
+\toctitle{Lecture Notes in Computer Science}
+\tocauthor{Authors' Instructions}
 \maketitle
+
 \begin{abstract}
-Abstraction providers sometimes need to directly introduce new types and operators into existing languages. 
-Ideally, such typed language fragments could be separately defined and safely composed. 
-Unfortunately, existing tools and techniques do not come equipped with modular reasoning principles comparable to those available for library-based embeddings packaged using a modern module system.
-Each combination of fragments is semantically its own dialect of the language and metatheoretic and compiler correctness results must be established monolithically.
+Abstraction providers sometimes need to extend a programming language with new base types and their corresponding operators. 
+Ideally, such language fragments could be defined and reasoned about separately.
+Unfortunately, existing tools and techniques do not come equipped with strong modular reasoning principle -- fundamentally, %comparable to those available for library-based embeddings packaged using a modern module system.
+each combination of fragments becomes its own dialect of the language for which metatheoretic and compiler correctness results must be established monolithically. 
 Recent work has started to address this problem, e.g.
-by showing how to modularly reason about concrete syntax.
-Our focus here is on safely composing separately defined type system fragments.
-We organize each fragment around a type constructor, as is usual practice when describing type systems, and sidestep the difficulties of abstract syntax by statically delegating semantic control over a small, fixed abstract syntax in a type-directed manner to static logic associated with a relevant type constructor.
-The paper is organized around a minimal calculus, @$\lambda$ that permits surprisingly practical examples.
+by showing how to modularly reason about desugarings atop a fixed type system.
+Our focus here is on safely composing type system fragments.
+We organize fragments around type constructors, as is usual practice when describing type systems, and sidestep the difficulties of composing abstract syntax (i.e.
+the expression problem)  by delegating control over a small, uniform abstract syntax in a type-directed manner to logic associated with a type constructor.
+The paper is organized around a minimal calculus, @$\lambda$, that permits surprisingly practical examples.
 We establish several strong semantic guarantees, notably \emph{type safety}, \emph{stability of typing} under extension and \emph{conservativity}:
@@
 This involves lifting typed compilation techniques into the semantics and enforcing an abstraction barrier around each fragment using a form of ``internal'' type abstraction. 
-Conservativity then follows from classical parametricity results, so these \emph{modular type constructors} can be reasoned about like modules.
-%We discuss how to optionally mechanize this reasoning in Coq.%We call embeddings constructed using this strengthened formulation of active type constructors \emph{active embeddings}.
-A strength of the technique is that it does not require mechanized specifications and proofs to make the guarantees above, but these can be constructed, using existing proof assistants, to increase confidence that an active type constructor definition adequately implements a specified fragment.%We introduce @$\lambda$, a simply-typed lambda calculus with simply-kinded type-level computation where new indexed type and operator constructors can be declared  from within.
-%It is structured in four layers: at the top-level, extension providers declare new indexed type constructors and term-level operator constructors.
-%Type-level functions associated with operator constructors define their static and dynamic semantics, the latter by translation to a fixed typed nternal language.
-By lifting compiler correctness techniques into the language, the ``actively typed semantics'' guarantees type safety.
-Going further, the semantics enforce an abstraction barrier at extension boundaries that ensures that extensions are mutually conservative (i.e.
-they do not weaken or interfere with one another, so that they can  be used together in any combination).
-%%Totality of the type-level language guarantees decidability of type checking and equality constraints on type indices makes type equivalence decidable. 
-%We intend @$\lambda$ as a minimal foundation for future work on safe language-integrated extension mechanisms for typed programming languages, but it is already quite expressive.
-We demonstrate by showing how a conventional concrete syntax can be introduced by type-directed dispatch to Core @$\lambda$, then discuss a number of typed language fragments that can be \emph{actively embedded} by this mechanism as orthogonal ``libraries''.
-%(and discuss the sorts of fragments that, as yet, cannot).
+Conservativity then follows from classical parametricity results, so these \emph{modular type constructors} can be reasoned about like modules in an ML-like language. 
+%\keywords{extensible languages; typed compilation; type-level computation; type abstraction}
 \end{abstract}
 
-%\category{D.3.2}{Programming Languages}{Language Classifications}[Extensible Languages]
-%\category{D.3.4}{Programming Languages}{Processors}[Compilers]
-%\category{F.3.1}{Logics \& Meanings of Programs}{Specifying and Verifying and Reasoning about Programs}[Specification Techniques]
-%\keywords
-%type-level computation, typed compilation
 \section{Introduction}
@@
 
-\section{@$\lambda$, By Example}\label{overview}
-
-\begin{figure}
-\small
-\hspace{-5px}\input{syntax-table-IL}
-\caption{Syntax of {$\mathcal{L}\{{\rightharpoonup}\,{\forall}\,{\mu}\,{\iunit}\,{\times}\,{+}\}$}, our internal language (IL).
-We write internal types and terms in blue when they may contain the indicated ``unquote'' forms, discussed later.}
-\label{syntax-IL}
-%\vspace{-10px}
-\end{figure}
-
-Programs in @$\lambda$ are written as \emph{external terms}, $e$.
-The concrete syntax of external terms is essentially conventional, as suggested by the left side of the example in Figure \ref{example}.
-It desugars (purely syntactically) to a subtantially more uniform abstract syntax, shown in Figure \ref{syntax-EL} and discussed as we go on.
-We review recent techniques that permit modularly introducing new desugarings like these in Sec.
-\ref{desugaring}.
-
-The static and dynamic semantics are specified simultaneously by a \emph{bidirectionally typed translation semantics} targeting an \emph{internal language}, with terms $\iota$.
-The two key judgements have the form:  \[\esynX{e}{\st\moutput}{\iota\moutput} ~~~~~\text{and}~~~~~ \eanaX{e}{\st}{\iota\moutput}\]
-\noindent
-
-These are pronounced ``$e$ (synthesizes / analyzes against) type $\sigma$ with translation $\iota$ under typing context $\Upsilon$ and tycon context $\Phi$''.
-We indicate ``outputs'' when introducing judgement forms by a \emph{mode annotation}, shown; these are not part of the judgement's syntax.
-The rules, shown in Figure \ref{statics-EL}, will be discussed incrementally as we explain the example in Figure \ref{example}.
-Bidirectional typechecking is also sometimes called \emph{local type inference} \cite{Pierce:2000:LTI:345099.345100}.
-%We will see how we leverage this bidirectionality when we discuss literal forms below.
-% The judgement is only well-defined for \emph{valid} tycon contexts, written judgementally as $\vdash \Phi$ and \emph{valid} typing contexts, $\vdash_\Phi \Upsilon$, which we also specify below.
- \begin{figure*}[t!]
-\small
-\hspace{-5px}\input{example-table}
-%\\~\\
-%$\begin{array}{rcl}
-%e\cdot\conclbl{lbl}(\conclbl{lbl}_1=e_1, \cdots, \conclbl{lbl}_n=e_n)  :=  \etarg{(\conclbl{lbl}; \svar{list}n[\klbl]~\conclbl{lbl}_1~\cdots~\conclbl{lbl}_n)}{e}{e_1; \cdots; e_n}
-%\end{array}$
-\caption{An example external term, $e_{ex}$, in concrete syntax (left), desugared to abstract syntax (right), with static terms shown in green in examples (only).
-The notation ${\small \scolor{\desugar{...}}}$ stands for an embedding of the indicated ${\small \scolor{\conclbl{label}}}$, ${\small \scolor{\concrx{regular expression}}}$, ${\small \scolor{\concstr{string}}}$ or numeral into the SL, with derived kinds  $\klbl$, $\krx$,  $\kstr$ and $\knat$, not shown.
-The signatures of helper functions, e.g.
-${\small \scolor{\svar{nil}}}$ and ${\small \scolor{\svar{list}n}}$, are shown in Figure \ref{helper-sigs}.}
-\label{example}
-\end{figure*}
-
-\paragraph{Internal Language} @$\lambda$ relies on a \emph{typed internal language} with support for type abstraction.
-In this paper, we use {$\mathcal{L}\{{\rightharpoonup}\,{\forall}\,{\mu}\,{\iunit}\,{\times}\,{+}\}$} (Figure \ref{syntax-IL}), though our results do not depend on this precise choice of internal type constructors, only on the general forms of the judgements.
-We assume the internal statics are specified in the standard way by judgements for internal type formation {$\iDelta~\iTheta \vdash \tau$}, typing context formation { $\iDelta \vdash \iGamma$} and type assignment {$\iDelta~\iGamma \vdash \iota : \tau\moutput$}.
-The internal dynamics are also a standard structural operational semantics with judgements {\small $\iota \mapsto \iota\moutput$} and {$\iota~\mathtt{val}$} (cf.
-\cite{pfpl}).
-
-We also here define a syntax for simultaneous substitutions, of terms for variables, $\gamma$, and of types for polymorphic type variables, $\delta$, and assume standard judgements ensuring that these are valid with respect to a valid context, $\Delta \vdash \gamma : \Gamma$ and $\vdash \delta : \Delta$.
-We apply these substitutions to terms, types and typing contexts using the syntax $[\gamma]\iota$, $[\delta]\iota$, $[\delta]\tau$ and $[\delta]\Gamma$ (in some prior work, application of a substitution like this is written using a hatted form, e.g.
-$\hat\gamma(\iota)$; we intend the same semantics but use a notation more consistent with standard substitution, e.g.
-$[\iota/x]\iota'$).
-We will omit  leading $\emptyset$ and $\cdot$ in examples; in specifications, the former is used for  metatheoretic finite mappings, the latter for metatheoretic ordered lists.
-
-This style of specification has some  similarities to the Harper-Stone \emph{typed elaboration semantics} for Standard ML \cite{Harper00atype-theoretic}.
-There, however, external and internal terms were governed by a common type system.
-In @$\lambda$, {internal types} classify only internal terms.
-This style may thus also be compared to specifications for the first stage of a type-directed compiler, e.g.
-the TIL compiler for Standard ML \cite{tarditi+:til-OLD}, here lifted ``one level up'' into the language.
-We will see how the techniques developed to reason about typed compilation are lifted into our semantics to compositionally reason about type safety, which requires that the translation of every well-typed external term is a well-typed internal term, below.
-
-\paragraph{Kinds and Static Terms}
-\begin{figure}[t!]
-\small
-\hspace{-5px}\input{syntax-table-EL}\\
-\caption{Abstract syntax of the external language (EL). Terms in gray are technical devices with no corresponding concrete syntax.}
-\label{syntax-EL}
-\end{figure}
-\begin{figure}[t!]
-\small
-\hspace{-4px}\input{syntax-table-SL}
-\caption{Syntax of the static language (SL).}
-\label{syntax-SL}
-\end{figure}
-The workhorse of our calculus is the \emph{static language}, which itself forms a typed lambda calculus where 
-%External terms are classified by (external) \emph{types}. 
-\emph{kinds}, $\kappa$, classify \emph{static terms}, $\sigma$.
-The syntax of the SL is given in Figure \ref{syntax-SL}.
-We note that the core of the SL is a total functional programming language based on several standard fragments: total functions, quantification over kinds, inductive kinds (constrained by a positivity condition to prevent non-termination), and products and sums (these closely follow \cite{pfpl} so we again omit some details).
-
-The kinding judgement, specified  in Figure \ref{statics-SL}, takes the form $\sofkX{\st}{\kappa\moutput}$, where $\kDelta$ and $\kGamma$ are analagous to $\iDelta$ and $\iGamma$ (and analagous well-formedness judgements are defined, also omitted).
-The natural number $n$ is used as a technical device in our metatheory to prevent terms of the form $\sana{n}{\st}$ and $\ssyn{n}$ from arising if not permitted; these forms, discussed later, never need to be written directly (they would have no corresponding concrete syntax), so $n$ can assumed 0 for all user code.
-$\kXi$ is described below.
-The normalization judgement (specified in Figure \ref{dynamics-SL}) takes the form $\snorm{\st}{\memD}{\memG}{\aCtx}{\st\moutput}{\memD\moutput}{\memG\moutput}$, where $\memD$, $\memG$ and $\aCtx$ are also technical devices that will be described later; they too can be ignored from the perspective of user code.
-We write $\st \Downarrow \st'$ iff $\snorm{\st}{\emptyset}{\cdot}{{\rightharpoonup}; \emptyset; \emptyset}{\st'}{\emptyset}{\cdot}$.
-\emph{Static values} are normalized static terms.
-Normalization can also raise an error (to indicate a type error in an external term, or a problem in a tycon definition, as we will discuss), indicated by the judgement $\serrX{\st}$.
-We omit error propagation rules.%We will refer to the relevant rules as we proceed.
-
- %We will specify the statics (i.e.
-kinding) and dynamics (i.e.
-normalization) of the static language in Figures \ref{statics-SL} and \ref{dynamics-SL}.%The normalization judgement takes the form $\snorm{\st}{\memD}{\memG}{\aCtx}{\st\moutput}{\memD\moutput}{\memG\moutput}$.
-%This core can be considered  a simply-typed dialect of a language like Coq (we explore this further in Sec.
-\ref{coq}).
-
-\paragraph{External Types are Static Values}
-
-External types (or simply \emph{types}) are {static values} of kind $\kty$.  %We will return to kinds $\kity$ and $\kitm$ below.
-The form $\sty{c}{\st}$ introduces a type by ``applying'' a \emph{tycon}, $c$, to a static term, $\st$, called the \emph{type index}.
-There is one built-in tycon, $\rightharpoonup$, classifying possibly partial external functions, discussed below.
-All other types are constructed by tycons defined in the tycon context and identified by name.
-We write tycon names in small caps, e.g.
-$\tcvar{lprod}$, and use $\tc$ as a metavariable ranging over these (Figure \ref{syntax-TC}).
-We will return to the other introductory form for types, $\sotherty{m}{\tau}$, which also serves only as a technical device, later.
-%\todo{mention otherty here?}
-
-Two user-defined types are constructed on line 1 of Figure \ref{example}: a labeled product type declaring a single field labeled $\conclbl{venue}$ of type $\tcvar{rstr}~\concrx{([A-Z]+) \digit\{4\}}$, which classifies \emph{regular strings}, known statically to be in the regular language of the indicated regular expression, which can contain nested parenthesized groups.
-%We write these indices using conventional concrete syntax.
-%The regular expression can contain parenthesized groups.% We will discuss the operations these provide below.
-%Although the concrete syntax resembles function application, in our calculus tycons are not themselves static terms.
-
-\paragraph{Index Kinds} The rule \rulename{k-ty} in Figure \ref{statics-SL} requires that the type index must be of the \emph{index kind} of the tycon, given by the \emph{index kind context}, $\kXi$, a simple mapping from tycons to kinds (Figure \ref{syntax-TC}).
-The initial index kind context, $\kXi_0$, is always included, defining the index kind of ${\rightharpoonup}$ as $\kty \times \kty$.
-The example index kind context in Figure \ref{example-tcsigs}, $\kXi_\text{ex}$, specifies that the index kind of $\tcvar{lprod}$ is $\klist{\kprod{\klbl}{\kty}}$, and the  index kind of $\tcvar{rstr}$ is $\krx$.
-We assume $\klist{\kappa}$, $\klbl$ and $\krx$ are embedded into the static language in the standard way.
-The concrete syntax used on line 1 desugars to static terms of these kinds (this sugar could be defined modularly as \emph{kind-specific syntax} by adapting the technique described in Sec.
-\ref{desugaring} \cite{TSLs}).
-If a tycon context, $\Phi$, is valid, it  uniquely determines a valid index kind context, as specified  in Figure \ref{statics-TC} by the judgement $\tccok{\Phi}{\kXi\moutput}$, detailed below.
-In our example, $\tccok{\Phi_\text{ex}}{\kXi_\text{ex}}$.
-
-\paragraph{Type-Level Computation}
-The labeled product type on line 1 is bound to the static variable $\svar{V}$  in the remainder of the external term.
-This  desugars to the more general external construct $\eslet{\st}{\svar{x}}{e}$, on the right.
-The rules \rulename{syn-slet} and \rulename{ana-slet}, shown in Figure \ref{statics-EL}  (here, the former applies) begin by assigning a kind to $\sigma$ under the index kind context determined by $\Phi$, then substitute its value into the external term $e$ to continue with synthesis or analysis.
-Static variables, written in bold, are distinct from external and internal variables.
-The rule \rulename{n-ty} in Figure \ref{dynamics-SL} shows that normalizing $\sty{c}{\st}$ simply involves normalizing $\st$.
-Because types are static values but need not be written directly in normal form, this is  a form of \emph{type-level computation of higher kind}, though our SL has a more general role than a standard type-level language.
-%Type indices (e.g.
-common regular expressions) and, as we will shortly discuss, term indices can also be computed.%, so our use of a static language type-level computation.
-
-\begin{figure}[t]
-\small
-\hspace{-5px}$\begin{array}{rcllcl}
-\multicolumn{3}{l}{\mvlbl{tycons}~~~c~::=~{\rightharpoonup}\,~|~ \tc} & \multicolumn{3}{l}{\mvlbl{index kind contexts}}\\
-\multicolumn{3}{l}{\mvlbl{tycon contexts}} & \kXi_0 & := & {\rightharpoonup}[\kty \times \kty]\\
-\Phi & ::= & \cdot ~|~ \Phi, \phi & \kXi & ::= & \kXi_0 ~|~ \kXi, \tc[\kappa]\\
-\multicolumn{3}{l}{\mvlbl{tycon defs}} & \multicolumn{3}{l}{\mvlbl{tycon sigs}}\\
-\phi & ::= & \tcdef{\tc}{\psi}{\opdefs}~~~~& \psi & ::= & \tcsig{\kappa}{\opsigs}\\
-\multicolumn{3}{l}{\mvlbl{opcon structures}} & \multicolumn{3}{l}{\mvlbl{opcon sigs}}\\
-\opdefs & ::= & \keyw{rep} = \st  & \Omega & ::= & \cdot\\
- & ~|~ & \omega, \analitop{\st} & & ~|~ & \Omega, \litsig{\kappa}\\
- & ~|~ & \opdefs, \synop{\opname{op}}{\st} & & ~|~ & \opsigs, \opsig{\opname{op}}{\kappa}
-\end{array}$
-\caption{Syntax of tycons and opcons + sigs and contexts.}
-\label{syntax-TC}
-\end{figure}
-%
-\begin{figure}
-\small
-\hspace{-5px}$\begin{array}{l@{\hspace{2px}}c@{\hspace{2px}}l@{\hspace{10px}}l@{\hspace{2px}}c@{\hspace{2px}}l}
-\Phi_\text{ex} & := & \phi_\text{rstr}, \phi_\text{lprod} & \kXi_\text{ex} & := &\kXi_0, \tcvar{rstr}[\krx], \tcvar{lprod}[\klist{\klbl\times\kty}]\\
-\phi_\text{rstr} & & \text{\normalsize(Figure \ref{tc-rstr})} & \psi_\text{rstr} & := &\tcsig{\krx}{\Omega_\text{rstr}}\\
-\phi_\text{lprod}& &\text{\normalsize(Figure \ref{tc-lprod})} & \psi_\text{lprod} &:=& \tcsig{\klist{\klbl \times \kty}}{\Omega_\text{lprod}}\\
-\Omega_\text{rstr} & := & \multicolumn{4}{r}{\litsig{\kstr}, \opsig{\opname{concat}}{\kunit}, \opsig{\opname{\#}}{\keyw{Nat}}, \opsig{\opname{coerce}}{\krx}, \opsig{\opname{case}}{\kunit}, \opsig{\opname{!rc}}{\kunit}}\\
-\Omega_\text{lprod} & := & \multicolumn{4}{r}{\litsig{\klist{\klbl}}, \opsig{\opname{\#}}{\klbl}, \opsig{\opname{with}}{\klist{\klbl}}, \opsig{\opname{!ri}}{\kunit}}
-\end{array}$
-\caption{The contexts and signatures for Figures \ref{tc-rstr} and \ref{tc-lprod}.}
-\label{example-tcsigs}
-\end{figure}
-
-\paragraph{Type Constructor Contexts}
-
-\begin{figure}[t]
-\small
-~\hfill{\fbox{$\tccok{\Phi}{\kXi\moutput}$}}\vspace{-20px}
-\begin{mathpar}
-\hspace{-45px}\inferrule[tcc-emp]{ }{\tccok{\cdot}{\kXi_0}}
-~~~~~~~~~~~~
-\inferrule[tcc-ext]{
-  \tccok{\Phi}{\kXi}\\
-  \tcok{\kXi}{\phi}{\tc}{\ktyidx}
-}{\tccok{\Phi, \phi}{\kXi, \tc[\ktyidx]}}
-\end{mathpar}
-%
-%
-~\hfill{\fbox{$\tcok{\kXi}{\phi}{\tc\moutput}{\kappa\moutput}$}}\vspace{-20px}
-%
-%
-\begin{mathpar}
-\hspace{-65px}\inferrule[tcdef-ok]{
-  \tc \notin \text{dom}(\kXi)~~~~
-  \tcsigok{\psi}{\ktyidx}~~~~
-  \ocok{\kXi}{\omega}{\tc}{\psi}
-}{
-  \tcok{\kXi}{\tcdef{\tc}{\psi}{\omega}}{\tc}{\ktyidx}
-}
-\end{mathpar}
-~\hfill{\fbox{$\tcsigok{\psi}{\kappa\moutput}$}}\vspace{-20px}
-\begin{mathpar}
-\hspace{-50px}\inferrule[tcsig-ok]{
-  \keq{\emptyset}{\ktyidx}\\
-  \ocsigok{\Omega}
-}{
-  \tcsigok{\tcsig{\ktyidx}{\Omega}}{\ktyidx}
-}
-\end{mathpar}
-~\hfill\fbox{$\keq{\kTheta\minput}{\kappa\minput}$}\vspace{-16px}\begin{mathpar}
-\hspace{-50px}\inferrule[keq-k]{\kvar \in \kTheta}{\keq{\kTheta}{\kvar}}
-~~~~~~
-\inferrule[keq-ind]{\keq{\kTheta, \kvar}{\kappa}}{
-  \keq{\kTheta}{\kmu{\kvar}{\kappa}}}
-~~~~~~
-\inferrule[keq-unit]{ }{\keq{\kTheta}{\kunit}}
-\\
-\inferrule[keq-prod]{
-  \keq{\kTheta}{\kappa_1}~~~~
-  \keq{\kTheta}{\kappa_2}
-}{
-  \keq{\kTheta}{\kprod{\kappa_1}{\kappa_2}}
-}
-~~~~~~
-\inferrule[keq-sum]{
-  \keq{\kTheta}{\kappa_1}~~~~
-  \keq{\kTheta}{\kappa_2}
-}{
-  \keq{\kTheta}{\ksum{\kappa_1}{\kappa_2}}
-}
-~~~~~~
-\inferrule[keq-ty]{ }{\keq{\kTheta}{\kty}}
-\end{mathpar}
-~\hfill\fbox{$\ocsigok{\Omega}$}\vspace{-18px}
-\begin{mathpar}
-\hspace{-20px}\inferrule[ocs-rep]{ }{\ocsigok{\cdot}}
-~~~~
-\inferrule[ocs-lit]{
-  \keq{\emptyset}{\kappa}
-}{
-  \ocsigok{\litsig{\kappa}}
-}
-~~~~
-\inferrule[ocs-targ]{
-  \ocsigok{\Omega}~~~~
-  \opname{op} \notin \text{dom}(\Omega)~~~~
-  \keq{\emptyset}{\kappa}
-}{
-  \ocsigok{\Omega, \opsig{\opname{op}}{\kappa}}
-}
-\end{mathpar}
-$\karg := (\kunit \rightarrow (\kty \times \kitm)) \times (\kty \rightarrow \kitm)$~\hfill\fbox{$\ocok{\kXi}{\omega}{\tc}{\psi}$}%\vspace{-18px}
-\begin{mathpar}
-\inferrule[oc-rep]{
-  \sofkz{\emptyset}{\emptyset}{\kXi}{\stx{rep}}{\karrow{\ktyidx}{\kity}}
-}{
-  \ocok{\kXi}{\keyw{rep}=\stx{rep}}{\tc}{\tcsig{\ktyidx}{\cdot}}
-}
-
-\inferrule[oc-lit]{
-  \ocok{\kXi}{\omega}{\tc}{\tcsig{\ktyidx}{\Omega}}\\
-  \sofkz{\emptyset}{\emptyset}{\kXi, \tc[\ktyidx]}{\stx{def}}{{\ktyidx}\rightarrow{\kappa}\rightarrow{\kargs}\rightarrow {\kitm}}
-}{
-  \ocok{\kXi}{\omega, \analitop{\stx{def}}}{\tc}{\tcsig{\ktyidx}{\Omega, \litsig{\kappa}}}
-}
-
-\inferrule[oc-targ]{
-  \ocok{\kXi}{\omega}{\tc}{\tcsig{\ktyidx}{\Omega}}\\
-  \sofkz{\emptyset}{\emptyset}{\kXi, \tc[\ktyidx]}{\stx{def}}{{\ktyidx}\rightarrow\kitm\rightarrow{\kappa}\rightarrow{\kargs}\rightarrow{(\kty\times \kitm)}}\\
-}{
-  \ocok{\kXi}{\omega, \synop{\opname{op}}{\stx{def}}}{\tc}{\tcsig{\ktyidx}{\Omega, \opsig{\opname{op}}{\kappa}}}
-}
-\end{mathpar}
-\caption{Type constructor kinding.}
-\label{statics-TC}
-\end{figure} A tycon context, $\Phi$, is a list of tycon definitions, $\phi$ (Figure \ref{syntax-TC}).
-Each tycon definition is annotated with a \emph{tycon signature}, $\psi$, which specifies the index kind of the tycon as well as an \emph{opcon signature}, $\Omega$, which relates to the \emph{opcon structure}, $\omega$, discussed below.
-The tycons used in our example are defined in Figures \ref{tc-rstr} and \ref{tc-lprod}.
-In Figure \ref{statics-TC}, the judgement $\tcok{\kXi}{\phi}{\tc\moutput}{\ktyidx\moutput}$ ensures that their names are unique, their signatures are valid, and that the opcon structure respects the opcon signature.%, ``returning'' the information needed to extend $\kXi$.
-
-% \begin{figure}[t]
-% \small
-% $\begin{array}{rcl}
-% \karg & := & (\kunit \rightarrow (\kty \times \kitm)) \times (\kty \rightarrow \kitm)\\
-% %\svar{arity0} : \klist{\karg} \rightarrow \kunit & \svar{syn} : \karg \rightarrow \kty \times \kitm\\
-% %\svar{arity1} : \klist{\karg} \rightarrow \karg & \svar{ana} : \karg \rightarrow \kty \rightarrow \kitm\\
-% %\svar{arity2} : \klist{\karg} \rightarrow \karg \times \karg
-% \end{array}$
-% \caption{The external semantics exposes arguments as list of pairs of static functions for synthesis and analysis, respectively. We assume the two helper functions above in the examples.}
-% \label{karg}
-% \end{figure}
-\begin{figure}[t]
-\small
-%\renewcommand{\slam}[3]{\lambda #2{:}#1.#3}
-\hspace{-5px}$\begin{array}{l}
-\tcdef{\tcvar{rstr}}{\psi_\text{rstr}}{\\
-\quad\keyw{rep}=\scolorc\slam{\krx}{\svar{tyidx}}{\sqity{\qcolorc\imu{s}{\keyw{str}\times\keyw{list}[s]}\scolorc}\color{black}}\\
-% ana lit
-\quad \analitop{
-  \scolorc\slam{\krx}{\svar{tyidx}}{\slam{\kstr}{\svar{tmidx}}{\slam{\kargs}{\svar{args}}{\cnl
-\quad\quad \keyw{let}~\svar{aok} : \kunit = \svar{arity0}~\svar{args}~\keyw{in}~\cnl
-\quad\quad \keyw{letpair}~(\svar{chars}, \svar{groups})  = \svar{rmatch}~\svar{tyidx}~\svar{tmidx}~\keyw{in}~\cnl
-\quad\quad\quad \sqitm{\qcolorc\ifold{s}{\keyw{str}\times\keyw{list}[s]}{\cnl\qcolorc
-\quad\quad\quad\quad (\quq{\scolorc\svar{str\_of\_Str}~\svar{chars}\qcolorc},\quq{\scolorc\svar{rstrs\_of\_Strs}~\svar{groups}\qcolorc})}\scolorc}}}}
-}\color{black},\\
-% syn concat
-\quad \synop{\opname{concat}}{\scolorc\slam{\krx}{\svar{ttyidx}}{\lambda \svar{ttr}{:}\kitm.\slam{\kunit}{\svar{tmidx}}{\slam{\kargs}{\svar{args}}{\cnl
-\quad\quad \keyw{letpair}~(\svar{targ}, \svar{a})  =\svar{arity2}~\svar{args}~\keyw{in}~\keyw{letpair}~(\svar{aty}, \svar{atr})=\svar{syn}~\svar{a}~\keyw{in}\cnl
-\quad\quad \stycase{\tcvar{rstr}}{\svar{aty}}{\svar{atyidx}}{(\sty{\tcvar{rstr}}{\svar{rconcat}~\svar{ttyidx}~\svar{atyidx}},\cnl
-\quad\quad\quad \sqitm{\qcolorc rsconcat~\quq{\scolorc\svar{ttr}\qcolorc}~\quq{\scolorc\svar{atr}\qcolorc}\scolorc})}{\sraise{\kty \times \kitm}}}}}}\color{black},\\
-% syn case
-%\quad \synop{\opname{case}}{\slam{\krx}{\svar{ttyidx}}{\slam{\kitm}{\svar{ttr}}{\slam{\kunit}{\svar{tmidx}}{\slam{\kargs}{\svar{args}}{\\
-%\quad\quad \svar{arity2}~\svar{args}~(\slam{\karg}{\svar{a1}}{\slam{\karg}{\svar{a2}}{\svar{syn}~\svar{a1}~\slam{\kty}{\svar{a1ty}}{\slam{\kitm}{\svar{a1tr}}{\\
-%\quad\quad\quad \keyw{let}~\svar{hdty} : \kty =\sty{\tcvar{rstr}}{\svar{rxhd}~\svar{ttyidx}}~\keyw{in}\\
-%\quad\quad\quad \keyw{let}~\svar{tlty} : \kty =\sty{\tcvar{rstr}}{\svar{rxtl}~\svar{ttyidx}}~\keyw{in}\\
-%\quad\quad\quad \keyw{let}~\svar{a2ty} : \kty =\sparr{(\svar{hdty}, \sparr{(\svar{tlty}, \svar{a1ty})})}~\keyw{in}\\
-%\quad\quad\quad \svar{ana}~\svar{a2}~\svar{a2ty}~\slam{\kitm}{\svar{a2tr}}{
-%        (\svar{a1ty}, \\
-%\quad\quad\quad\quad \sqitm{\iAp{\qtuq{\srep{\svar{a1ty}}}}{scase}~\quq{\svar{ttr}}~\ilam{\iunit}{\_}{\quq{\svar{a1tr}}}~\quq{\svar{a2tr}}})}}}}}}}}}}\\
-%% syn eq
-%\quad \synop{\opname{eq}}{\slam{\krx}{\svar{ttyidx}}{\slam{\kitm}{\svar{ttr}}{\slam{\kunit}{\svar{tmidx}}{\slam{\kargs}{\svar{args}}{\\
-%\quad\quad \svar{arity3}~\svar{args}~\slam{\karg}{\svar{a1}}{\slam{\karg}{\svar{a2}}{\slam{\karg}{\svar{a3}}{\\
-%\quad\quad\quad \svar{syn}~\svar{a1}~\slam{\kty}{\svar{a1ty}}{\slam{\kitm}{\svar{a1tr}}{\stycase{\tcvar{rx}}{\svar{a1ty}}{\svar{a1tyidx}}{\\
-%\quad\quad\quad \svar{synsame}~\svar{a2}~\svar{a3}~\slam{\kty}{\svar{bty}}{\slam{\kitm}{\svar{a2tr}}{\slam{\kitm}{\svar{a3tr}}{(\svar{bty},\\
-%\quad\quad\quad\quad \sqitm{\iAp{\qtuq{\srep{\svar{bty}}}}{sequal}~\quq{\svar{ttr}}~\quq{\svar{a1tr}}~\\
-%\quad\quad\quad\quad\quad \ilam{\iunit}{\_}{\quq{\svar{a2tr}}}~\ilam{\iunit}{\_}{\svar{a3tr}}})}}}}{err}}}}}}}}}}}\\
-% syn grp
-\quad \synop{\opname{\#}}{\scolorc\slam{\krx}{\svar{ttyidx}}{\slam{\kitm}{\svar{ttr}}{\slam{\knat}{\svar{tmidx}}{\slam{\kargs}{\svar{args}}{\cnl
-\quad\quad \keyw{let}~\svar{aok} : \kunit =\svar{arity0}~\svar{args}~\keyw{in}\cnl
-\quad\quad \keyw{let}~\svar{rg} : \krx = \svar{rgroupn}~\svar{ttyidx}~\svar{tmidx}~~\keyw{in}\cnl
-\quad\quad\quad (\sty{\tcvar{rstr}}{\svar{rg}}, \sqitm{\qcolorc rsgroupn~\quq{\scolorc\svar{nat\_of\_Nat}~\svar{tmidx}\qcolorc}~\quq{\scolorc\svar{ttr}\qcolorc}\scolorc})
-}}}}}\color{black},\\
-% syn coerce
-\quad \synop{\opname{coerce}}{\scolorc\slam{\krx}{\svar{ttyidx}}{\slam{\kitm}{\svar{ttr}}{\slam{\krx}{\svar{tmidx}}{\slam{\kargs}{\svar{args}}{\cnl
-\quad\quad \keyw{let}~\svar{aok} : \kunit = \svar{arity0}~\svar{args}~\keyw{in}\cnl
-\quad\quad \keyw{let}~\svar{slok} : \kunit = \svar{rsublang}~\svar{tmidx}~\svar{ttyidx}~\keyw{in}\cnl
-\quad\quad \keyw{let}~\svar{rtr} : \kitm = \svar{rx\_of\_Rx}~\svar{tmidx}~\keyw{in}\cnl
-\quad\quad \quad (\sty{\tcvar{rstr}}{\svar{tmidx}},\sqitm{\qcolorc rsregroup~\quq{\scolorc\svar{rtr}\qcolorc}~\quq{\scolorc\svar{ttr}\qcolorc}\scolorc})}}}}}\color{black},\\
-% syn check
-\quad \synop{\opname{case}}{\scolorc\slam{\krx}{\svar{ttyidx}}{\slam{\kitm}{\svar{ttr}}{\slam{\kunit}{\svar{tmidx}}{\slam{\kargs}{\svar{args}}{\cnl
-\quad\quad \keyw{letpair}~(\svar{a1}, \svar{a2})=\svar{arity2}~\svar{args}~\keyw{in}~\keyw{letpair}~(\svar{a1ty}, \svar{a1tr})=\svar{syn}~\svar{a1}~\keyw{in}~\cnl
-\quad\quad \sparrcase{\svar{a1ty}}{\svar{a1tyidx}}{\keyw{letpair} (\svar{a1tyin}, \svar{a1tyout}) = \svar{a1tyidx}~\keyw{in}\cnl
-\quad\quad \stycase{\tcvar{rstr}}{\svar{a1tyin}}{\svar{r}}{\keyw{let}~\svar{rtr}:\kitm = \svar{rx\_of\_Rx}~\svar{r}~\keyw{in}\cnl
-\quad\quad \keyw{let}~\svar{a2tr} : \kitm = \svar{ana}~\svar{a2}~\svar{a1tyout}~\keyw{in}\cnl
-\quad\quad\quad (\svar{a1tyout}, \sqitm{\qcolorc\iAp{\qtuq{\scolorc\srep{\svar{a1tyout}}\qcolorc}}{rscheck}~\quq{\scolorc\svar{rtr}\qcolorc}~\quq{\scolorc\svar{ttr}\qcolorc}~\cnl\qcolorc
-\quad\quad\quad\quad \quq{\scolorc\svar{a1tr}\qcolorc}~\ilam{\iunit}{\_}{\quq{\scolorc\svar{a2tr}\qcolorc}}\scolorc})\cnl
-\quad\quad}{\sraise{\kty \times \kitm}}}}{\sraise{\kty \times \kitm}}}}}}\\
-\quad \synop{\opname{!rc}}{\scolorc \lambda \svar{ttyidx}{:}\krx.\lambda \svar{ttr}{:}\kitm.\lambda \svar{tmidx}{:}{\kunit}.\lambda \svar{args}{:}\kargs.\cnl
-\quad\quad (\sqitm{\sty{\tcvar{rstr}}{\desugar{\concrx{\digit+}}}}, \svar{str\_of\_Str}~{\desugar{\concstr{13}}})}\color{black}}
-\end{array}
-$
-\caption{The definition of the $\tcvar{rstr}$ tycon, $\phi_\text{rstr}$.}
-\label{tc-rstr}
-\end{figure}
-\begin{figure}[t]
-\small
-\hspace{-5px}$\begin{array}{l}
-\tcdef{\tcvar{lprod}}{\psi_\text{lprod}}{\\
-\quad\keyw{rep}=\scolorc\lambda\svar{tyidx}{:}\klist{\klbl \times \kty}.\skap{\kity}{\skap{\klbl\times\kty}{\svar{listrec}}}~\svar{tyidx}~\sqity{\qcolorc\iunit\scolorc}~\cnl
-  \quad\quad (\lambda \svar{h}{:}\klbl \times \kty.\lambda \svar{r}{:}\kity.\keyw{letpair}~(\_,\svar{hty})=\svar{h}~\keyw{in}~\cnl
-  \quad\quad\quad \sqity{\qcolorc\qtuq{\scolorc\svar{r}\qcolorc}\times\qtuq{\scolorc\srep{\svar{hty}}\qcolorc}\scolorc})\color{black},\\
-\quad\analitop{\scolorc\lambda\svar{tyidx}{:}\klist{\klbl\times\kty}.\lambda\svar{tmidx}{:}\klist{\klbl}.\lambda\svar{args}{:}\kargs.\cnl
-  \quad\quad \keyw{let}~\svar{inhabited}:\kunit=\svar{uniqmap}~\svar{tyidx}~\keyw{in}\cnl
-  \quad\quad \skap{\kitm}{\skap{\karg}{\skap{\klbl}{\skap{\klbl\times\kty}{\svar{listrec3}}}}}}~\svar{tyidx}~\svar{tmidx}~\svar{args}~\sqitm{\qcolorc\itriv\scolorc}\cnl
-  \quad\quad\quad \lambda \svar{h1}{:}\klbl\times\kty.\lambda \svar{h2}{:}\klbl.\lambda\svar{h3}{:}\karg.\lambda \svar{r}{:}\kitm.\cnl
-  \quad\quad\quad\quad \keyw{letpair}~(\svar{rowlbl}, \svar{rowty})=\svar{h1}~\keyw{in}~\keyw{let}~\svar{lok}:\kunit=\svar{lbleq}~\svar{rowlbl} ~\svar{h2}~\keyw{in}\cnl
-  \quad\quad\quad\quad \keyw{let}~\svar{rowtm} : \kitm = \svar{ana}~\svar{h3}~\svar{rowty}~\keyw{in}~\sqitm{\qcolorc(\quq{\scolorc\svar{r}\qcolorc}, \quq{\scolorc\svar{rwtm}\qcolorc})\scolorc}\color{black},\\
-\quad\synop{\opname{\#}}{\scolorc\lambda \svar{ttyidx}{:}\klist{\klbl\times\kty}.\lambda \svar{ttr}{:}\kitm.\lambda\svar{tmidx}{:}\klbl.\lambda\svar{args}{:}\kargs.\cnl
-  \quad\quad \keyw{let}~\svar{aok} : \kunit = \svar{arity0}~\svar{args}~\keyw{in}\cnl
-  \quad\quad \keyw{letpair}~(\svar{rownum}, \svar{rowty})=\svar{lookup}~\svar{ttyidx}~\svar{tmidx}~\keyw{in}\cnl
-  \quad\quad\quad (\svar{rowty}, \svar{prjnth}~\svar{rownum}~\svar{ttr}))}\color{black},\\
-\quad\synop{\opname{with}}{\scolorc\cnl
-\quad\lambda \svar{ttyidx}{:}\klist{\klbl\times\kty}.\lambda \svar{ttr}{:}\kitm.\lambda\svar{tmidx}{:}\klist{\klbl}.\lambda\svar{args}{:}\kargs.\cnl
-  \quad\quad \keyw{letpair}~(\svar{olist}, \svar{otr})=\skap{\klist{\klbl\times\kty}\times\kitm}{\skap{\karg}{\skap{\klbl}{\svar{listrec2}}}}~\cnl
-  \quad\quad\quad \svar{tmidx}~\svar{args}~(\svar{ttyidx}, \svar{ttr})\cnl
-  \quad\quad\quad (\lambda \svar{h1}{:}{\klbl}.\lambda\svar{h2}{:}{\karg}.\lambda\svar{r}{:}{\klist{\klbl\times\kty}\times\kitm}.\cnl
-  \quad\quad\quad\quad \keyw{letpair}~(\svar{rowty}, \svar{rowtr})=\svar{syn}~\svar{h2}~\keyw{in}~\keyw{letpair}~(\svar{rlist}, \svar{rtr})=\svar{r}~\keyw{in}\cnl
-  \quad\quad\quad\quad\quad (\skap{\klbl\times\kty}{\svar{append}}~\svar{rlist}~(\svar{h1}, \svar{rowty}),\cnl
-  \quad\quad\quad\quad\quad \sqitm{\qcolorc(\quq{\scolorc\svar{rtr}\qcolorc}, \quq{\scolorc\svar{rowtr}\qcolorc})\scolorc}))~\keyw{in}~\cnl
-  \quad\quad \keyw{let}~\svar{inhabited}:\kunit=\svar{uniqmap}~\svar{olist}~\keyw{in}~\cnl
-  \quad\quad (\sty{\tcvar{lprod}}{\svar{olist}}, \svar{otr})}\\
-\quad\synop{\opname{!ri}}{\scolorc \lambda \svar{ttyidx}{:}\krx.\lambda \svar{ttr}{:}\kitm.\lambda \svar{tmidx}{:}{\kunit}.\lambda \svar{args}{:}\kargs.\cnl
-\quad\quad (\sqitm{\sty{\tcvar{rstr}}{\desugar{\concrx{\digit+}}}}, \sqitm{\qcolorc\ifold{s}{\keyw{str}\times\keyw{list}[s]}{\\
-\quad\quad\quad \qcolorc(\quq{\scolorc\svar{str\_of\_Str}~\desugar{\concstr{oops}}\qcolorc},\quq{\scolorc\svar{rstrs\_of\_Strs}~\skap{\kstr}{\svar{nil}}\qcolorc})}\scolorc})}\color{black}}
-\end{array}$
-\caption{The definition of the $\tcvar{lprod}$ tycon, $\phi_\text{lprod}$.}
-\label{tc-lprod}
-\end{figure}
-\paragraph{Type Equivalence} 
-To simplify the handling of type equivalence, type index kinds must be \emph{equality kinds}: those for which semantic equivalence coincides with syntactic equality at normal form.
-We define these by the judgement $\keq{\kDelta}{\kappa}$, appearing as a premise of \rulename{tcsig-ok} in Figure \ref{statics-TC}.
-Equality kinds are similar to equality types as found in Standard ML.
-The main implication of this choice is that type indices cannot contain static functions%, an issue we will return to in Sec.
-\ref{variants} when discussing polymorphic and recursive  types.
-%Type equality is not an extensible part of the calculus, an issue that we will return to below.
-%We implicitly identify terms up to alpha renaming of bound variables throughout this paper, as is the usual convention (algorithmically, a locally nameless representation can be used).
-
-\paragraph{Subsumption} The rule \rulename{subsume}, Figure \ref{statics-EL}, simply states that if a term synthesizes a type, it can be analyzed against that type.
-This relies on the fact that equivalent types are syntactically identical, and a metatheoretic property that states that the type synthesis judgement, assuming well-formed contexts, actually synthesizes a type (stated formally in  Section \ref{calculus}).
-Subsumption will only be applied once in our example, on line 7, discussed below.
-
-\paragraph{Representations}
-\begin{figure}
-\small
-\small
-~\hfill\fbox{$\sfinalrepX{\st}{\tau\moutput}$}\vspace{-18px}
-\begin{mathpar}
-\inferrule[conc-rep]{
-  \boxdot = \text{names}(\Phi)\\\\
-  \snorm{\srep{\st}}{\emptyset}{\cdot}{\boxdot,{\rightharpoonup}; \emptyset; \Phi}{\sqity{\ity}}{\memD}{\cdot}\\
-  \vdash_\Phi \memD \leadsto \delta : \Delta\\
-  \Delta~\emptyset \vdash \tau
-}{
-  \sfinalrep{\Phi}{\st}{[\delta]\tau}
-}
-\end{mathpar}
-~\hfill\fbox{$\sfinalrepX{\Upsilon}{\Gamma\moutput}$}\vspace{-18px}
-\begin{mathpar}
-\hspace{-75px}\inferrule[trans-ctx-emp]{ }{
-  \sfinalrep{\Phi}{\emptyset}{\emptyset}
-}
-~~~~~~~~
-\inferrule[trans-ctx-ext]{
-  \sfinalrep{\Phi}{\Upsilon}{\Gamma}\\
-  \sfinalrep{\Phi}{\st}{\tau}
-}{
-  \sfinalrep{\Phi}{\Upsilon, x \Rightarrow \st}{\Gamma, x : \tau}
-}
-\end{mathpar}
-~\hfill\fbox{$\vdash_\Phi \memD \leadsto \delta\moutput : \Delta\moutput$}\vspace{-18px}
-\begin{mathpar}
-\hspace{-74px}\inferrule[D-emp]{ }{\emptyset \leadsto \emptyset : \emptyset}
-~~~~~
-\inferrule[D-ext]{
-  \vdash_\Phi \memD \leadsto \delta : \Delta\\
-  \vdash_\Phi \st \leadsto \tau
-}{
-  \vdash_\Phi \memD, \st \leftrightsquigarrow \alpha \leadsto \delta, \tau/\alpha : \Delta, \alpha
-}
-\end{mathpar}
-\caption{Concrete Representations}
-\label{representations}
-\end{figure}
-Each external type has a corresponding internal type called its \emph{concrete representation}.
-Each tycon definition specifies the concrete representations of the types it constructs with a \emph{representation schema}, specified as the first component of the opcon structure.
-Rule \rulename{oc-rep} specifies that this is simply a static function mapping type indices to static terms of kind $\kity$, which are introduced by the quotation form $\sqity{\qcolorc\tau\color{black}}$.
-Unquote forms, written $\qcolorc\qtuq{\color{black}\st\qcolorc}$, normalize away  recursively, shown in rules \rulename{n-qity-*}.
-
-In Figure \ref{tc-rstr}, the representation schema of $\tcvar{rstr}$ simply ignores the type index, mapping all regular string types to the same internal type: a recursive type pairing a string  with a list of other regular strings, corresponding to the ``pre-extracted'' parenthesized groups (internal types $\keyw{str}$ and $\keyw{list}[\tau]$ are defined in the usual way, and are distinguished from analagous kinds $\keyw{Str}$ and $\keyw{List}$ by initial capitalization).
-This is of course not the only valid choice of representation -- one could simply use $\keyw{str}$ (increasing the cost of group extraction, discussed below), or use a tree of offsets, or inspect the index to extract the number of groups, permitting the use of nested tuples (in practice, the IL might provide fixed length vectors).
-%This should remind readers of the situation of choosing different representation types for an abstract type declared my a module signature.
-As we will see, the internal representation of a type constructed in this way is similarly held abstract outside of the opcon structure, so this choice is up to the tycon definition.
-
-In Figure 9, the representation schema of $\tcvar{lprod}$ constructs a simple nested tuple representation based on the type index (the labels are not needed in the translation).
-Because the type index itself contains types, the representation schema must refer to the representations of these types.
-The static language provides the operator $\srep{\st}$ to extract knowledge about the representation of type $\st$.
-The kinding rule \rulename{k-rep} is simple, but the normalization semantics are more interesting.
-Rule \rulename{n-rep-conc}  extracts the representation schema and  passes it the type index, but it does not always apply.
-For types constructed by a tycon other than ones named in the \emph{concrete rep whitelist}, $\boxdot$ (a simple list of tycons), an \emph{abstract representation}, is generated.
-The first time a representation is requested, a fresh internal type variable, $\alpha$, is generated and the correspondence is recorded in the \emph{abstract rep store}, $\memD$, as seen in rule \rulename{n-rep-abs}.
-Later requests return this type variable.
-Requesting the representation is a simple effect.
-Given $\svar{V}$ from our example and abbreviating the index of the type of the field $\conclbl{venue}$ as $\stx{rsidx}$: \[\scriptsize\snorm{\srep{\normalsize\svar{V}\scriptsize}}{\emptyset}{\cdot}{{\rightharpoonup},\tcvar{lprod}; \emptyset; \Phi_\text{ex}}{\sqity{\iunit\times\alpha}}{\emptyset, \sty{\tcvar{rstr}}{\stx{rsidx}} \leftrightsquigarrow \alpha}{\cdot}\]
-
-The judgement $\vdash_\Phi \memD \leadsto \delta\moutput : \Delta\moutput$ constructs a substitution and corresponding internal type variable context by generating the concrete representations for each stored type.
-For the above example: \[\vdash_{\Phi_\text{ex}} \emptyset, \sty{\tcvar{rstr}}{\stx{rsidx}} \leftrightsquigarrow \alpha \leadsto \emptyset, \imu{s}{\keyw{str} \times \keyw{list}[s]}/\alpha : \emptyset, \alpha\]
-
-%We will see how this underlies our results in Sec. \ref{calculus}.
-The judgement $\sfinalrepX{\st}{\tau\moutput}$ extracts the concrete representation by placing every tycon in $\Phi$ in the whitelist and checking that the representation is well-formed (${\rightharpoonup}$ is always in the whitelist; $\memD$ may still be non-empty because of $\sotherty{m}{\tau}$, also discussed later):\[\snorm{\srep{\svar{V}}}{\emptyset}{\cdot}{{\rightharpoonup},\tcvar{rstr},\tcvar{lprod}; \emptyset; \Phi_\text{ex}}{\sqity{\iunit\times\imu{s}{\keyw{str}\times\keyw{list}[s]}}}{\emptyset}{\cdot}\]% The representations of the types in Figure \ref{example} need not be known to understand its semantics.
-
-\paragraph{Variables}
-
-\begin{figure*}[t]
-\small
-~\hfill\fbox{$\esynX{e}{\st\moutput}{\iota\moutput}$}
-~\fbox{$\eanaX{e}{\st}{\iota\moutput}$}
-\begin{mathpar}
-\inferrule[subsume]{
-  \esynX{e}{\st}{\iota}
-}{
-  \eanaX{e}{\st}{\iota}
-}
-
-\inferrule[syn-slet]{
-  \tccok{\Phi}{\kXi}\\
-  \sofkz{\emptyset}{\emptyset}{\kXi}{\st}{\kappa}\\\\
-  \snormT{\st}{\st'}\\
-  \esynX{[\st'/\sx]e}{\stx{ty}}{\iota}
-}{
-  \esynX{\eslet{\st}{\sx}{e}}{\stx{ty}}{\iota}
-}
-
-\inferrule[ana-slet]{
-  \tccok{\Phi}{\kXi}\\
-  \sofkz{\emptyset}{\emptyset}{\kXi}{\st}{\kappa}\\\\
-  \snormT{\st}{\st'}\\
-  \eanaX{[\st'/\sx]e}{\stx{ty}}{\iota}
-}{
-  \eanaX{\eslet{\st}{\sx}{e}}{\stx{ty}}{\iota}
-}
-
-\inferrule[ascribe]{
-  \tccok{\Phi}{\kXi}\\
-  \sofkz{\emptyset}{\emptyset}{\kXi}{\st}{\kty}\\\\
-  \snormT{\st}{\st'}\\
-  \eanaX{e}{\st'}{\iota}
-}{
-  \esynX{\easc{e}{\st}}{\st'}{\iota}
-}
-
-\inferrule[syn-var]{
-  x \Rightarrow \st \in \Upsilon
-}{
-  \esynX{x}{\st}{x}
-}
-
-\inferrule[syn-let]{
-  \esynX{e_1}{\st_1}{\iota_1}\\\sfinalrepX{\st_1}{\tau_1}\\\\
-  \esyn{\Upsilon, x \Rightarrow \st_1}{\Phi}{e_2}{\st_2}{\iota_2}\\
-}{
-  \esynX{\elet{e_1}{x}{e_2}}{\st_2}{\iap{\ilam{\tau_1}{x}{\iota_2}}{\iota_1}}
-}
-
-\inferrule[ana-let]{
-  \esynX{e_1}{\st_1}{\iota_1}\\\sfinalrepX{\st_1}{\tau_1}\\\\
-  \eana{\Upsilon, x \Rightarrow \st_1}{\Phi}{e_2}{\st_2}{\iota_2}\\
-}{\eanaX{\elet{e_1}{x}{e_2}}{\st_2}{\iap{\ilam{\tau_1}{x}{\iota_2}}{\iota_1}}}
-
-\inferrule[ana-fix]{
-  \sfinalrepX{\st}{\tau}\\\\
-  \eana{\Upsilon, x \Rightarrow \st}{\Phi}{e}{\st}{\iota}
-}{
-  \eanaX{\efix{x}{e}}{\st}{\ifix{\tau}{x}{\iota}}
-}
-
-\inferrule[syn-lam]{
-  \tccok{\Phi}{\kXi}\\
-  \sofkz{\emptyset}{\emptyset}{\kXi}{\st_1}{\kty}\\
-  \snormT{\st_1}{\st_1'}\\
-  \sfinalrepX{\st_1'}{\tau_1}\\\\
-  \esyn{\Upsilon, x \Rightarrow \st_1'}{\Phi}{e}{\st_2}{\iota}
-}{
-  \esynX{\elam{\st_1}{x}{e}}{\sparr{(\st_1',\st_2)}}{\ilam{\tau_1}{x}{\iota}}
-}
-
-\inferrule[syn-ap]{
-  \esynX{e_1}{\sparr{(\st_1, \st_2)}}{\iota_1}\\\\
-  \eanaX{e_2}{\st_2}{\iota_2}
-}{
-  \esynX{\eap{e_1}{e_2}}{\st_2}{\iap{\iota_1}{\iota_2}}
-}
-
-\inferrule[ana-lit]{
-  \tcdef{\tc}{\tcsig{\ktyidx}{\Omega}}{\omega} \in \Phi\\\\
-  \litsig{\klitidx} \in \Omega\\
-  \tccok{\Phi}{\kXi}\\
-  \sofkz{\emptyset}{\emptyset}{\kXi}{\stmidx}{\klitidx}\\\\
-  \analitop{\stx{def}} \in \omega\\
-  \mkargsX\\\\
-  \snorm{\stx{def}~\sttyidx~\stmidx~\stx{args}}{\emptyset}{\memG_0}{{\rightharpoonup}, \tc; \Upsilon; \Phi}{\sqitm{\itm_\text{abs}}}{\memD}{\memG}\\\\
-  \snorm{\srep{\sty{\tc}{\sttyidx}}}{\memD}{\cdot}{{\rightharpoonup}, \tc;\emptyset;\Phi}{\sqity{\ity_\text{abs}}}{\memD'}{\cdot}\\\\
-  \vdash_\Phi \memD' \leadsto {\delta} : \iDelta\\
-  \memG \leadsto {\gamma} : \iGamma\\
-  \iDelta~\iGamma \vdash \iota_\text{abs} : \tau_\text{abs}
-}{
-  \eanaX{\elit{\stmidx}{\es}}{\sty{\tc}{\sttyidx}}{[{\delta}][{\gamma}]\iota_\text{abs}}
-}
-
-\inferrule[syn-targ]{
-  \esynX{e_\text{targ}}{\sty{\tc}{\sttyidx}}{\iota_\text{targ}}\\\\
-  \tcdef{\tc}{\tcsig{\ktyidx}{\Omega}}{\omega} \in \Phi\\\\
-  \opsig{\opname{op}}{\ktargidx} \in \Omega\\
-  \tccok{\Phi}{\kXi}\\
-  \sofkz{\emptyset}{\emptyset}{\kXi}{\stmidx}{\ktargidx}\\\\
-  \synop{\opname{op}}{\stx{def}} \in \omega\\
-  \mkargs{e_\text{targ}; \es}{\stx{args}}{\memG_0}{n}\\\\
-  %\snorm{\srep{\sty{\tc}{\sttyidx}}}{\emptyset}{\cdot}{{\rightharpoonup}, \tc;\emptyset;\Phi}{\sqity{\ity_\text{targ}}}{\memD_0}{\cdot}\\\\
-  \snorm{\stx{def}~\sttyidx~(\svar{tr}~\ssyn{0})~\stmidx~(\svar{atl}~\stx{args})}{\emptyset}{\memG_0}{{\rightharpoonup},\tc; \Upsilon; \Phi}{(\st, \sqitm{\itm_\text{abs}})}{\memD}{\memG}\\\\
-  \snorm{\srep{\st}}{\memD}{\cdot}{{\rightharpoonup},\tc;\emptyset;\Phi}{\sqity{\ity_\text{abs}}}{\memD'}{\cdot}\\\\
-  \vdash_\Phi \memD' \leadsto {\delta} : \iDelta\\
-  \memG \leadsto {\gamma} : \iGamma\\
-  \iDelta~\iGamma \vdash \iota_\text{abs} : \tau_\text{abs}
-}{
-  \esynX{\etarg{\opname{op}}{\sttmidx}{e_\text{targ}}{\es}}{\st}{[\delta][\gamma]\iota_\text{abs}}
-}
-
-\inferrule[ana-other]{
-  \sfinalrep{\Phi}{\srep{\Upsilon}}{\Gamma \parallel \memD}\\
-  \memD \leadsto \delta; \Delta\\
-  \Delta~\Gamma \vdash \iota : \tau
-}{
-  \eanaX{\eother{\iota}}{\sotherty{n}{\tau}}{[\delta]\iota}
-}
-\end{mathpar}
-\caption{Typing}
-\label{statics-EL}
-\end{figure*}
-
-We can now continue to line 2 of Figure \ref{example}, where we bind a term to a variable using the form $\elet{e_1}{x}{e_2}$.
-The rules \rulename{syn-let} and \rulename{ana-let} give an essentially standard semantics.
-The translation is to lambda application.
-The concrete representation of the synthesized type for $e_1$ determines the type annotation.
-The \rulename{syn-var} rule is standard.
-The well-formedness judgement for external typing contexts, $\ctxok{\Phi}{\Upsilon}$, ensures that variables map to types (Figure \ref{statics-EL2}).
-The judgement $\vdash_\Phi \Upsilon \leadsto \Gamma\moutput$ translates the types in $\Upsilon$ to their concrete representations (Figure \ref{representations}).
-
-%The rule \rulename{ana-fix} similarly permits fixpoint definitions in analytic positions (this fixpoint operator follows Plotkin's PCF, cf. \cite{pfpl}). 
-\paragraph{Functions}
-
-On lines 4-6, we see a function definition.
-The syntax and semantics of functions are fixed and built into the external language (so that the semantics can uniformly handle the tricky issue of variable binding).
-Only the argument types need to be provided explicitly; the return type is synthesized, as shown in the rules \rulename{syn-lam}.
-Rule \rulename{syn-ap} is standard.
-Function types are written concretely as $\sigma_1 \rightharpoonup \sigma_2$ and abstractly as $\sparr{(\sigma_1, \sigma_2)}$, i.e.
-the index kind of the built-in type constructor $\rightharpoonup$ is $\kprod{\kty}{\kty}$ (cf.
-$\kXi_0$ in Figure \ref{example-tcsigs}).
-By the semantics of the operators described below, the full type synthesized by $paper$, written concretely, will be:
-\[\begin{array}{l}
-\concty{rstr}{\concrx{..*}} \rightharpoonup \concty{rstr}{\concrx{\digit\digit*}} \rightharpoonup \concty{labeledprod}{\{\\
-\quad \conclbl{venue} : \concty{rstr}{\concrx{([A-Z]+) \digit+}}, \conclbl{title} : \concty{rstr}{\concrx{..*}},\\
-\quad \conclbl{doi} : \concty{rstr}{\concrx{\digit\digit\rdot\digit\digit\digit\digit\textbackslash/\digit\digit*}}\}}
-\end{array}\]
- Like internal functions (and unlike static functions), external functions may be partial, via a $\keyw{fix}$point operator.
-Rule \rulename{ana-fix} follows Plotkin's PCF, cf.
-\cite{pfpl}, here defined analytically.
-%(an analytic lambda could also be defined).
-
-\paragraph{Ascriptions} Type ascriptions allow the programmer to explicitly specify the type a term should be analyzed against.
-For analytic terms appearing in synthetic positions, an ascription is necessary.
-In the abstract syntax, type ascriptions are always placed directly on terms using the form $\easc{e}{\sigma}$ (in the concrete syntax, they may appear on the variable being bound for convenience).
-The rule \rulename{syn-asc} in Figure \ref{statics-EL} specifies that the ascription must be a closed static term of kind $\kty$ under the index kind context determined by the tycon context.
-It is normalized to a type before proceeding with analysis.
-This is also an essentially standard rule.
-
-\paragraph{Literals} The variable $v$ binds a term introduced \emph{literally}, under an ascription specifying that it should be analyzed against type $\svar{V}$.
-Inside this literal, we also see a literal, though with no ascription (it will be analyzed against the type specified in $\svar{V}$, as discussed below).
-All literal forms in the concrete syntax desugar to the same abstract form, $\elit{\stmidx}{\es}$.
-The \emph{term index},  $\stmidx$, captures statically known portions of the literal as a single static term and the \emph{argument list}, $\es$,  captures all external sub-terms.
-For example, in the labeled product literal, desugaring constructs a list of field labels as the term index (using standard helper functions whose signatures are shown in Figure \ref{helper-sigs}) and passes in the  corresponding field values as arguments.
-For regular string literals, the string is lifted into the SL as the term index.
-There are no sub-terms, so the argument list is empty.
-
-Literal terms can only appear in analytic positions in our calculus. More specifically, they are given meaning by {the type constructor of the type they are being analyzed against}. 
-This tycon's opcon signature, $\Omega$,  determines the \emph{literal term index kind} governing $\stmidx$.
-In Figure \ref{example-tcsigs}, $\psi_\text{lprod}$ specifies $\Omega_\text{lprod}$ which in turn specifies the literal term index kind $\klist{\klbl}$, and $\psi_\text{rstr}$ similarly specifies $\Omega_\text{rstr}$ and ${\kstr}$.
-The literal index kind must also be an equality kind, as specified by rule \rulename{ocs-lit} in Figure \ref{statics-TC}.
-The rule governing type analysis of literals is \rulename{ana-lit}, shown in Figure \ref{statics-EL}.
-It begins by checking the provided term index against this kind (premises 1-4).% Note that here, the index kind context tracks only type index kinds, as operators are not themselves constructed as static values \todo{mention future work thing?}.
-
-\subparagraph{Literal Opcon Definitions}
-
-The next premise of \rulename{ana-lit} extracts the \emph{literal opcon definition}, $\stx{def}$, defined using the form $\analitop{\stx{def}}$ in the opcon structure, $\omega$, of the tycon definition.
-This  is responsible for deciding the translation of the literal as a function of the type index, the term index and  the types and translations of the arguments.
-%The judgement $\ocok{\kXi}{\omega}{\tc}{\psi}$, in Figure \ref{statics-TC}, checks the opcon definition list $\omega$ for tycon $\tc$ against its signature $\psi$.
-
-The relevant rule in Figure \ref{statics-TC} is \rulename{oc-lit}, which ensurses that $\stx{def}$ has kind $\ktyidx \rightarrow \klitidx \rightarrow \kargs \rightarrow \kitm$.
-The kinds $\ktyidx$ and $\klitidx$ are determined by the signature as just described.
-The kind $\kitm$ is analagous to $\kity$, and has one introductory ``quotation'' form, $\sqitm{\qcolor{\iota}}$.
-Normalization produces an internal term without such forms (rules \rulename{n-qitm-*} in Figure \ref{dynamics-SL}), i.e.
-a standard IL term.
-Note that these can only be opaquely composed, not inspected (there is no elim form) and $\kity$ and $\kitm$ are not equality kinds.
-
-Before discussing arguments and the remaining premises, let us turn to the literal opcon definition for $\tcvar{rstr}$ in Figure \ref{tc-rstr}, which will be called by the rule \rulename{ana-lit} to derive a translation for the field value on line 1, $e_\text{venue} := \elit{\desugar{\concstr{EXMPL 2015}}}{\cdot}$: \footnotesize\[\begin{array}{l}\eana{\emptyset}{\Phi_\text{ex}}{e_\text{venue}}{\sty{\tcvar{rstr}}{\desugar{\concrx{([A-Z]+) \digit+}}}}{\ifold{s}{\keyw{str}\times\keyw{list}[s]}{\\
-\quad \keyw{pair}(\llfloor\concstr{EXMPL 2015}\rrfloor; \iAp{\imu{s}{\keyw{str}\times\keyw{list}[s]}}{list1}~\ifold{s}{\keyw{str}\times\keyw{list}[s]}{\\
-\quad\quad \keyw{pair}(\llfloor\concstr{EXMPL}\rrfloor; \iAp{\imu{s}{\keyw{str}\times\keyw{list}[s]}}{nil})})}}
-\end{array}\]
-%where we define the representation \footnotesize$\tau_{rs} := \imu{s}{\keyw{s} \times \keyw{list}[s]}$
-\normalsize
-We write $\llfloor{...}\rrfloor$ for the embedding of the indicated string into the IL, and abbreviate the definitions of $\keyw{str}$, $\keyw{list}[s]$ and the helper functions for working with lists as above.
-We refer to the translation as $\iota_\text{venue}$.
-
-The logic in Figure \ref{tc-rstr} simply checks that the argument list is empty using the simple helper function $\svar{arity0}$, which raises an error if the list is non-empty (rules \rulename{k-raise} and \rulename{n-raise}; in practice, the provider would here provide an explicit error message to the client).
-It then calls the helper function $\svar{rmatch}$ on the term index to make sure it matches the regular expression provided as the type index, also extracting its groups statically.
-Finally, it constructs the translation above by calling simple helper functions for lowering static strings and lists into the IL using quotations.
-We do not show the details; this is, by design, exactly the code one would write in a compiler written in a simply-typed functional language for the branch of the case analysis in the translation logic corresponding to the regular string literal term constructor.
-Here, we have pulled it into the tycon definition, pushed the case analysis into the semantics, and do not require a separate term constructor by relying on bidirectional typechecking to permit us to share syntax amongst many types.% (see Sec.
-\ref{desugaring}).%Reasoning about the correctness of this function requires specifying stronger invariants than the kind system provides, and our purpose in this section is not to claim to guarantee extension correctness.
-Our point is that mistakes will be caught
-
-\paragraph{Arguments} Let us now return to the important issue of how the semantics handles arguments by discussing how the literal opcon definition for $\tcvar{lprod}$ in Figure \ref{tc-lprod} is called by \rulename{ana-lit} to derive the following translation (assuming the definition of $\svar{list1}$):
-\footnotesize\[\begin{array}{l}
-\stx{tmidx} := \skap{\klbl}{\svar{list1}}~\desugar{\conclbl{venue}}\\
-\stx{tyidx} := \conckap{\svar{list1}}{\kprod{\klbl}{\kty}}~(\desugar{\conclbl{venue}}, \sty{\tcvar{rstr}}{\stx{rsidx}})\\
-\eana{\emptyset}{\Phi_\text{ex}}{\elit{\stx{tmidx}}{e_\text{venue}}}{\sty{\tcvar{lprod}}{\stx{tyidx}}}{\ipair{\itriv}{\iota_\text{venue}}}
-\end{array}\]
-\normalsize
-Note that we did not here exploit the opportunity to optimize the representation of labeled products of length 1.
-Our interest here is not in this particular decision,  but to ensure that such decisions can be freely explored and have only local implications, like changes in the representation of an abstract type in an ML-like module system.
-
-To call the opcon definition, the semantics needs to construct the \emph{argument interface}, $\stx{args}$, a list of type $\klist{\karg}$.
-%It also constructs an \emph{argument interface list}, $\stx{args}$ as well as an \emph{argument store}, $\memG_0$, from the argument list, $\es$, as specified by the judgement $\mkargsX$ in Figure \ref{statics-EL2}.
-The judgement also reports the lengths of these, $n$.
-The kind $\karg$ is defined in Figure \ref{statics-TC}: it classifies a pair of static functions that serve as hooks into the semantics, permitting the opcon definition to request synthesis or analysis, respectively, for the argument these functions abstract.
-The judgement $\mkargs{\es}{\st\moutput}{\memG\moutput}{n\moutput}$, specified in Figure \ref{statics-EL2}, constructs a list of such function pairs by wrapping the static operators $\ssyn{n}$ and $\sana{n}{\st}$.
-The kinding rules \rulename{k-syn} and \rulename{k-ana} (and the signatures in Figure \ref{helper-sigs}) imply that $\sofkn{\emptyset}{\emptyset}{\kXi}{n}{\stx{args}}{\kargs}$, where $n$ is the number of arguments.
-All user-defined static terms, as we have seen, are always checked with $n=0$, ensuring that these operators cannot arise directly in well-kinded opcon definitions,  avoiding the possibility of ``out of bounds'' problems.
-Passing the argument interface into opcon definitions, which were checked with $n=0$, does not cause kind safety problems due to a simple lemma, provable inductively:
-\begin{lemma}
-If $\sofkn{\kDelta}{\kGamma}{\kXi}{n}{\st}{\kappa}$ and $n' > n$ then $\sofkn{\kDelta}{\kGamma}{\kXi}{n'}{\st}{\kappa}$.
-\end{lemma}
-
-The arguments are placed in an \emph{argument store}, $\memG$.
-Initially, $\memG_0$ is simply a numbered sequence of the arguments themselves, here $\memG_0 := \cdot, 0 \hookrightarrow e_\text{venue}$.
-The normalization semantics for $\ssyn{n}$ and $\sana{n}{\st}$ can have an effect on the argument store.
-
-The literal opcon constructor for $\tcvar{lprod}$ in Figure \ref{tc-lprod} begins by ensuring that the type the literal is being analyzed against might actually be inhabited by ensuring that there are no duplicate labels in the type index.\footnote{Implementing a $\keyw{Map}[\kappa]$ kind that ensures this by construction, while remaining an equality kind, requires some simple additions to the SL, e.g.
-abstract equality kinds modeled also on SML, which we omit for concision and to demonstrate this more generally applicable technique.} It then recurses simultaneously over the term index, type index and argument list, checking corresponding labels for equality and requesting analysis of each argument against the corresponding type (the helper function $\svar{ana}$ is in Figure \ref{helper-sigs}; $\svar{listrec3}$ raises an error if the inputs are of different length).
-The rule \rulename{n-ana} shows how the argument is extracted from the store and analyzed in the typing context and tycon context provided (by rule \rulename{ana-lit} here) in the \emph{operation context}, $\aCtx$.
-If it succeeds, the argument store is updated with the type provided for analysis, the translation and, crucially, a fresh internal variable, $x$, standing for its translation.
-The abstract representation of the type provided for analysis is also produced using the whitelist in $\aCtx$, which contains only the tycon of the type the literal is being analyzed against (and the function tycon, which is never held abstract so functions can be used for operations with binding structure, e.g.
-\opname{case}, below).
-We write $\memG \otimes n \hookrightarrow ...$ to indicate that the previous value in the store for $n$ is removed.
-Because $\tcvar{rstr}$ is not in $\boxdot$, we will thus have:\footnotesize\[\begin{array}{l}
-\snorm{\stx{rep}~\sttmidx~\sttyidx~\stx{args}}{\emptyset}{\memG_0}{{\rightharpoonup},\tcvar{lprod}; \emptyset; \Phi_\text{ex}}{\sqitm{\ipair{\itriv}{x}}}{\memD}{\memG}\\
-\memD := \emptyset, \sty{\tcvar{rstr}}{\stx{rsidx}} \leftrightsquigarrow \alpha\\
-\memG := \cdot, 0 \hookrightarrow e_\text{venue} : \sty{\tcvar{rstr}}{\stx{rsidx}} \leadsto \iota_\text{venue}/x : \alpha
-\end{array}
-\]\normalsize
-
-\paragraph{Abstract Translation Checking} We must now check that this \emph{abstract translation} is \emph{representationally consistent}.
-The next two premises in \rulename{ana-lit} generate the abstract representation of the type provided for analysis, as well as its corresponding substitution, $\delta$, and internal type variable context, $\Delta$, as was shown previously.
-The judgement $\memG \leadsto \gamma\moutput : \Gamma\moutput$ generates a term substitution and internal typing context based on $\memG$ similarly (Figure \ref{statics-EL2}).
-Here, $\memG \leadsto (\emptyset, \iota_\text{venue}/x) : (\emptyset, x : \alpha)$ is generated.
-Finally, the semantics checks the \emph{abstract translation} above against the abstract representation using these contexts, e.g.
-here $\emptyset, \alpha~\emptyset, x : \alpha \vdash \ipair{\itriv}{x} : \iunit \times \alpha$.
-It is easy to see that this holds.
-Only after performing this check are the substitutions $\delta$ and $\gamma$ applied.
-We will  show an example where this check fails even when the check would have succeeded if we had applied the substitutions first below  and return to the powerful metatheoretic implications of this check in Sec.
-\ref{calculus}.
-
-\paragraph{Hygiene} Note that the variables in $\Upsilon$ are not made available in $\Gamma$ when performing this check, guaranteeing \emph{hygienic translation}: the translation must not have any free variables other than those generated implicitly via the argument store (which applying the  substitution will eliminate) so inadvertent capture cannot occur.
-We assume also that substitution application is \emph{capture-avoiding} (i.e.
-the binding structure of terms is maintained, as can be implemented straightforwardly using a locally nameless representation).
-Thus, terms can be reasoned about without conditions on the context that translation logic may place them under, i.e.
-\emph{compositionally}.
-
-\paragraph{Targeted Operations}
-
-%To better motivate the normalization semantics of $\ssyn{n}$ and $\sana{n}{\st}$ in Figure \ref{dynamics-SL}, which update the argument store, let us look more closely at our examples.
-
-\begin{figure}
-~\hfill\fbox{$\ctxok{\Phi}{\Upsilon}$}\vspace{-18px}
-\begin{mathpar}
-\hspace{-35px}\inferrule[ectx-emp]{ }{
-  \ctxok{\Phi}{\cdot}
-}
-~~~~~~
-\inferrule[ectx-ext]{
-  \vdash \Phi : \kXi\\
-  \sofkz{\emptyset}{\emptyset}{\kXi}{\st}{\kty}\\
-  \snormT{\st}{\st}
-}{
-  \ctxok{\Phi}{\Upsilon, x \Rightarrow \st}
-}
-\end{mathpar}
-~\hfill\fbox{$\mkargs{\es}{\st\moutput}{\memG\moutput}{n\moutput}$}\vspace{-20px}
-\begin{mathpar}
-\hspace{-20px}\inferrule[mkargs-z]{
-  \snormT{\skap{\karg}{\svar{nil}}}{\stx{args}}
-}{
-  \mkargs{\cdot}{\stx{args}}{\cdot}{0}
-}
-
-\inferrule[mkargs-s]{
-  \mkargs{\es}{\stx{args}}{\memG}{n}\\
-  \snormT{\skap{\karg}{\svar{push}}~\stx{args}~((\lambda \svar{\_}{:}\kunit.\ssyn{n}), (\lambda \svar{t}{:}\kty.\sana{n}{\svar{t}}))}{\stx{args}'}
-}{
-  \mkargs{\es, e}{\stx{args}}{\memG, n \hookrightarrow e}{n+1}
-}
-\end{mathpar}
-~\hfill\fbox{$\memG \leadsto \gamma\moutput : \Gamma\moutput$}\vspace{-20px}
-\begin{mathpar}
-\hspace{-40px}\inferrule[G-emp]{ }{\cdot \leadsto \emptyset : \emptyset}
-~~~~~~~~
-\inferrule[G-ignored]{\memG \leadsto \Gamma}{
-  \memG, n \hookrightarrow e \leadsto \Gamma
-}
-
-\inferrule[G-ext]{
-  \memG \leadsto \gamma : \Gamma
-}{
-  (\memG, n \hookrightarrow e : \st \leadsto \iota/x : \tau) \leadsto (\gamma, \iota/x) : (\Gamma, x : \tau)
-}
-\end{mathpar}
-\caption{Auxiliary judgements for external statics.}
-\label{statics-EL2}
-\end{figure}
-
-On line 5 of Figure \ref{example}, we invoke the \emph{labeled product extension opcon}, $\opname{with}$, on $v$ to create a new labeled product with additional fields, $\conclbl{title}$ and $\conclbl{doi}$, with types consistent with those shown in the return type of $paper$ above.
-The $\conclbl{doi}$ field's value is computed by invoking the \emph{regular string concatenation opcon}, $\opname{concat}$.
-A number of other similar \emph{targeted operations} are seen elsewhere.
-These all desugar to the same abstract form, $\etarg{\opname{op}}{\stx{tmidx}}{e_\text{targ}}{\es}$, where metavariable $\opname{op}$ ranges over opcon names, $\stx{tmidx}$ again captures all statically known portions of the operation (e.g.
-the field names on line 5), $e_\text{targ}$ is the \emph{target} expression, and $\es$ contains all other external subterms.
-
-In our calculus, all targeted operations are synthetic.
-The type constructor of the type recursively synthesized for the target is delegated responsibility over the semantics of the targeted operation.
-It must define an \emph{opcon term index kind} in its opcon signature and an \emph{opcon definition} in its opcon structure matching the name provided, and satisfying the kinding conditions in rules \rulename{ocs-targ} and \rulename{oc-targ} of Figure \ref{statics-TC}.
-The rule \rulename{syn-targ} in Figure \ref{statics-EL} shows that type synthesis and translation of targeted operations shares much in common with the logic for literals just described.
-The term index is first checked against the term index kind, then the opcon definition is extracted and an argument interface is constructed as before.
-We include the target itself as the notional first argument, though for convenience we don't require that each opcon re-synthesize the translation by passing the components in as shown.
-Whereas literal opcon definitions only had to decide a translation, targeted opcon definitions must decide both a type and a translation, returning a pair of kind $\kty \times \kitm$.
-The abstract translation is checked against the abstract representation of the type produced, as  above.% Note that it might have a tycon other than the one delegated control.
-
-We will not describe each opcon in Figures \ref{tc-rstr} and \ref{tc-lprod} in detail.
-Once again, the code is essentially identical to the code one would write in a standard typechecker.
-The ``magic'' happens in the dynamics of the SL, only manifesting when there is a problem.
-We note the following interesting features in these examples:
-\begin{itemize}
-\item In \opname{concat}, the definition uses the elimination form for types, $\stycase{c}{\st}{\sx}{\st}{\st}$, to extract the type index, compute the concatenated regular expression and construct a translation that performs the concatenation.
-This implies that tycons in our calculus have features of both open sum types (a default case is required) and modules.
-Tycon indices are not held abstract.
-In the calculus as given, there is rarely a reason to inspect the index of a different tycon.
-If we added support for direct calls between opcons (e.g.
-so that the regular expression could expose an operator that returned a labeled product containing the groups directly), having this ability would become more useful.
-\item The \opname{coerce} operation supports converting between regular string types known statically to obey a sublanguage relation (cf.
-\cite{conf/cpp/DoczkalKS13}).
-Rearrangement of group boundaries can also be performed in this way (this requires a regrouping operation in this choice of translation).
-We invoke it on line 7 in an analytic position, so subsumption is applied.
-An interesting direction for future work is to add support for describing implicit coercions between types of the same tycon by enriching the subsumption rule.
-\item The \opname{case} operation supports checked conversions.
-The first argument must synthesize a function type, where the index of the input type determines the regular expression that the string will dynamically be checked against.
-The ``else branch'', which is necessary here but not for a coercion, is analyzed against the return type of this function.
-Note that on line 9, the else branch will never be taken, but the type system ``forgot'' the information that could be used to optimize it away.
-\item Both example tycons define a \opname{\#} opcon, though with different index kinds.
-They are seen being used on line 8 to extract a field and then a subgroup.
-Like literals, types are used to distinguish the semantics despite the shared syntax.
-This has similarities to operator overloading, but here the static and dynamic semantics themselves are being overloaded statically.
-%Subgroup extraction together with regrouping via coercion and case analysis obviates the need for more conventional elim forms.
-\item It is enlightening to derive the abstract translation and representation produced when typechecking $p\opname{\#}\conclbl{venue}$ on line 8, recalling that $p$ is a labeled product with three fields each of a different regular string type, with \conclbl{venue} being the first.
-Assuming \keyw{rs} is the concrete representation of regular strings: $\begin{array}{l}
-\iota_\text{abs} := \keyw{letpair}(x; \_, y.\keyw{letpair}(y; z, \_.z))\\
-\tau_\text{abs} := \alpha_1\\
-\delta := \emptyset, \keyw{rs}/\alpha_1, \keyw{rs}/\alpha_2, \keyw{rs}/\alpha_3\\
-\Delta := \emptyset, \alpha_1, \alpha_2, \alpha_3\\
-\gamma := \emptyset, p/x\\
-\Gamma := \emptyset, x : \iunit \times (\alpha_1 \times (\alpha_2 \times \alpha_3))
-\end{array}$
-\item The $\opname{!rc}$ opcon definition shows an example where the representational consistency check would fail simply because a term of a type other than the one determined by the representation schema of $\tcvar{rstr}$ was generated.
-The $\opname{!ri}$ opcon definition is more interesting, however: the translation produced matches the concrete type of $\tcvar{rstr}$, so type safety would not be violated if it were permitted and it would pass the check described above if it were performed post-substitution.
-But it is quite problematic for reasons of modular reasoning: it claims that the translation has a regular string type, but the string itself does not obey the \emph{value invariant} that $\tcvar{rstr}$ locally maintains: that the string in the translation is actually in the language specified by the type index.
-Luckily, it does not pass the abstract translation check because it is not \emph{representationally independent} ($\keyw{fold}$ cannot have type $\alpha$).
-\end{itemize}
-
-% \begin{figure}
-% ~\hfill\fbox{$\sfinalrepX{\srep{\Upsilon}}{\Gamma\moutput \parallel \memD\moutput}$}
-% \begin{mathpar}
-% \inferrule[rep-ctx-emp]{ }{
-%   \sfinalrep{\Phi}{\srep{\cdot}}{\cdot \parallel \cdot}
-% }
-
-% \inferrule[rep-ctx-ext]{
-%   \sfinalrepX{\srep{\Upsilon}}{\Gamma \parallel \memD}\\
-%   \snorm{\srep{\st}}{\memD}{\cdot}{{\rightharpoonup}; \emptyset; \Phi}{\sqity{\qity}}{\memD'}{\cdot}\\
-%   \qity \uparrow \tau
-% }{
-%   \sfinalrep{\Phi}{\srep{\Upsilon, x\Rightarrow\st}}{\Gamma, x : \tau \parallel \memD'}
-% }
-% \end{mathpar}
-% \caption{Do I need this?}
-% \end{figure}
-\section{Metatheory of @$\lambda$}\label{calculus}
-Let us now make these intuitions more rigorous.
-At the outset, a disclaimer: the {ideas} described in the previous section, unifying several disparate threads of research into an evidently powerful core calculus, are intended to serve as the main intellectual contributions of the paper.
-We can only present the theorems and lemmas below semi-formally.
-Proving the theorems below completely rigorously given the normalization semantics we have presented is difficult, because it often requires us to reason about  intermediate terms, rather than sub-terms.
-Rigorous proofs would be most straightforward if we had presented an equivalent structural operational semantics, but this would obscure some important aspects of our presentation.
-We hope to submit mechanized proofs using a more suitable specification style as an artifact, and have started, but not completed, this work using Coq as of submission.
-
- \paragraph{Type Safety} The first major issue that we must address is kind safety and type safety.
-Because the semantics of the EL relies on the kinding and normalization semantics of the SL, and the normalization semantics of the SL relies on the semantics of the EL, this is somewhat tricky.
-Let us define $\stype{\st}{\Phi}$ iff $\vdash \Phi \sim \kXi$ and $\svalofk{\st}{\kty}$ and $\snormT{\st}{\st}$. Then, we need: 
-\begin{theorem}[Representational Consistency]
-If $\tccokX$ and $\vdash_\Phi \Upsilon$ and $\vdash_\Phi \Upsilon \leadsto \Gamma$ then $\emptyset \vdash \Gamma$ and \begin{enumerate}
-\item If $\esynX{e}{\st}{\iota}$ then $\stypeX{\st}$ and $\vdash_\Phi \st \leadsto \tau$ and $\emptyset~\Gamma \vdash \iota : \tau$.
-\item If $\stypeX{\st}$ and $\eanaX{e}{\st}{\iota}$ then $\vdash_\Phi \st \leadsto \tau$ and    $\emptyset~\Gamma \vdash \iota : \tau$.
-\end{enumerate}
-\end{theorem}
-We must simultaneously prove several important lemmas to show Theorem 1.
-We assume suitable definitions of judgements $\vdash_\Phi \memD$, which makes sure that $\memD$ stores types, and $\memD~\Upsilon \vdash_\Phi^n \memG$, which ensure that the $n$ facts in $\memG$ are truly valid.
-% We can begin by observing that the size of the argument store never changes:
-
-% \begin{lemma}[Argument Store]
-% If $\tccokX$ and $\sofk{\emptyset}{\emptyset}{\kXi}{\st}{\kappa}$ and $\snormX{\st}{\st'}$ then $|\memG|=|\memG'|$.
-% \end{lemma}
-
-% We can next consider only normalization under an empty argument store, allowing us to exclude the $\keyw{ana}$ and $\keyw{syn}$ forms, which are not used in representation schemas, and prove the following mutually defined lemmas.
-
-Many of the cases require ensuring that the normalization semantics is well-behaved (we discuss canonical forms below):
-
-\begin{lemma}[Static Preservation and Normalization]
-If $\tccokX$ and $\sofk{\emptyset}{\emptyset}{\kXi}{\st}{\kappa}$  and $\vdash_\Phi \memD$ and $\memD~\Upsilon \vdash_\Phi^n \memG$  and $\snorm{\st}{\memD}{\memG}{\boxdot; \Upsilon; \Phi}{\st'}{\memD'}{\memG'}$then $\sofkz{\emptyset}{\emptyset}{\kXi}{\st'}{\kappa}$ and $\vdash_\Phi \memD'$ and $\memD'~\Upsilon \vdash_\Phi^n \memG'$ and $\snormT{\st'}{\st'}$.
-\end{lemma}
-We also need these lemmas about concrete representations.
-\begin{lemma}[Concrete Representations]
-If $\tccokX$ and $\stypeX{\st}$ and $\vdash_\Phi \st \leadsto \tau$ then $\emptyset~\emptyset \vdash \tau$.
-\end{lemma}
-\begin{lemma}[Ext. Typing Contexts]
-If $\tccokX$ and $\vdash_\Phi \Upsilon$ then \begin{enumerate}
-\item If $x \Rightarrow \st \in \Upsilon$ then $\stypeX{\st}$.
-\item If $\vdash_\Phi \Upsilon \leadsto \Gamma$ then $\emptyset \vdash \Gamma$.
-\end{enumerate}
-\end{lemma}
-Both these and the cases for literals and targeted forms rely on standard properties of substitutions:
-
-\begin{lemma}[Substitutions] If $\delta : \Delta$  and $\Delta \vdash \gamma : \Gamma$ and $\Delta~\Gamma \vdash \iota : \tau$, then $\emptyset \vdash [\delta]\Gamma$ and $\emptyset~\emptyset \vdash [\delta][\gamma]\iota : [\delta]\tau$, and if $\Delta~\emptyset \vdash \tau$ then $\emptyset~\emptyset \vdash [\delta]\tau$.
-\end{lemma} 
-We also need the assurance that normalization semantics are updating the store in a way that produces  valid substitutions.
-
-\begin{lemma}[Stores] If $\vdash \Phi \sim \kXi$ and $\vdash_\Phi \Upsilon$ and $\vdash_\Phi \memD$ and $\memD~\Upsilon \vdash_\Phi \memG$ and $\snorm{\st}{\memD}{\memG}{\boxdot; \Upsilon; \Phi}{\st'}{\memD'}{\memG'}$ then $\vdash_\Phi \memD' \leadsto \delta : \Delta$ and $\delta : \Delta$ and $\memG' \leadsto \gamma : \Gamma$ and $\Delta \vdash \gamma : \Gamma$.
-\end{lemma}
-
-Ensuring that induction is well-founded is the most complex part of these proofs.
-We observe that the representation schema for a tycon $\tc$ is kinded under a $\kXi$ that does not include $\tc$ (Figure \ref{statics-TC}).
-Thus, it cannot request the representation of a type constructed by $\tc$ unless that type is extracted from the index.
-The argument store used in typechecking a term contains only subterms and the normalization semantics only attempt to typecheck terms in the argument store.
-For static terms kinded with $n=0$, the argument store is left unchanged.
-In other circumstances, its domain is fixed.
-
-%\begin{lemma}[User Terms] If $\sofkz{\emptyset}{\emptyset}{\kXi}{\st}{\kappa}$ and $\snormX{\st}{\st'}$ then $\memG'=\memG$. \end{lemma}
-The representational consistency theorem is, in essence, a version of the \emph{type-preserving compilation theorem} developed for the TIL compiler for SML \cite{tarditi+:til-OLD}, here lifted into the language and using user-defined mappings from external types to internal types (governed by the lemmas proved above).
-It implies type safety (that well-typed programs cannot ``go wrong'') by simply setting $\Upsilon=\emptyset$ and invoking the internal type safety theorem:
-
-\begin{lemma}[Internal Type Safety]
-If $\emptyset~\emptyset \vdash \iota : \tau$ then either $\iota~\mathtt{val}$ or $\iota \mapsto \iota'$ such that $\emptyset~\emptyset \vdash \iota' : \tau$.
-\end{lemma}
-
-\paragraph{Unicity} It is straightforward to prove that the type and translation of a term is unique if it exists.
-Each form that does not simply defer to a sub-term is either analytic, or synthetic, never both.
-This is an important property, as it justifies our ``caching'' behavior for $\memG$ and ensures that typing is deterministic.
-
-\begin{theorem}[Unicity]
-If $\tccokX$ and $\vdash_\Phi \Upsilon$  and $\eanaX{e}{\st}{\iota}$ and $\eanaX{e}{\st'}{\iota'}$ then  $\st=\st'$ and $\iota=\iota'$.
-\end{theorem}
-
-Note that extending the calculus with analytic targeted operations (e.g.
-case analysis that does not require synthetic branches) needs care to ensure that unicity is maintained, due to subsumption (i.e.
-analysis can proceed either via an explicit analysis rule or via a synthesis rule).
-If we define judgements specifying that subsumption should only proceed if an explicit analysis rule fails, this can be addressed, at the expense of some awkwardness in the specification.
-
-\paragraph{Decidability} A useful property is \emph{decidability of typechecking}.
-Here, we have suggested a decision procedure using the mode annotations, and by ensuring that normalization cannot diverge.
-We discussed ensuring that requesting a representation, or requesting analysis and synthesis, will not diverge above.
-The equality kind condition on type indices subsumes the positivity condition that would be needed to ensure that the elim form for types does not introduce non-termination.
-We must, however, leave rigorously proving strong normalization of the calculus as future work.
-
-A subtle issue arises in rule \rulename{n-syn-fail} and \rulename{n-ana-fail}.
-In brackets, we assume a judgement defining the situation where typechecking fails, though we do not inductively define it.
-In practice, this would not be difficult to determine, though in the specification, it is awkward to write down these cases.
-If the bracketed premises are omitted, normalization becomes weak (i.e.
-non-deterministic), but only in that it may fail when it was possible for it to succeed.
-Similar issues arose in the Harper-Stone semantics for SML \cite{Harper00atype-theoretic}.
-
-\paragraph{Stability}
-We can now turn to the situation where the tycon context is extended. Doing so does not affect kinding or normalization.
-\begin{theorem}[Stable Kinding and Normalization]
-If $\vdash \Phi \sim \kXi$ and $\sofkX{\st}{\kappa}$ and $\vdash \Phi, \phi \sim \kXi, \tc[\kappa']$ then \begin{enumerate}
-\item $\sofk{\kDelta}{\kGamma}{\kXi, \tc[\kappa']}{\st}{\kappa}$.
-\item If $\snorm{\st}{\memD}{\memG}{\boxdot; \Upsilon; \Phi}{\st'}{\memD'}{\memG'}$ then $\snorm{\st}{\memD}{\memG}{\boxdot; \Upsilon; \Phi, \phi}{\st'}{\memD'}{\memG'}$.
-\end{enumerate}
-\end{theorem}
-
-Several other similar lemmas for judgements indexed by $\Phi$ are omitted.
-The consequence of this is that extending $\Phi$ does not affect previously well-typed terms.
-Note that type systems structured as ``bags of rules'', where extensions can pattern match arbitrarily, would not have this property simultaneously with unicity (i.e.
-there would be ambiguities).
-Our method of delegating to a tycon, which remains stable upon extension, is the key to maintaining stability.
-\begin{theorem}[Stable Typing]
-If $\vdash \Phi \sim \kXi$ and $\vdash_\Phi \Upsilon$ and $\vdash_\Phi \Upsilon \leadsto \Gamma$  and $\vdash \Phi, \phi \sim \kXi, \tc[\kappa']$ then
-\begin{enumerate}
-\item If $\esynX{e}{\st}{\iota}$ then $\esyn{\Upsilon}{\Phi, \phi}{e}{\st}{\iota}$.
-\item If $\eanaX{e}{\st}{\iota}$ then $\eana{\Upsilon}{\Phi, \phi}{e}{\st}{\iota}$.
-\end{enumerate}
-\end{theorem}
-
-\paragraph{Conservativity}
-We now turn to the issue of conservativity, i.e.
-ensuring that the new terms that become well-typed upon extension of the tycon context cannot violate type invariants maintained in the previous ``closed world''.
-We saw at the end of Sec.
-2 an example of a hazardous situation, where the $\tcvar{lprod}$ tycon attempted to introduce a term of type $\sty{\tcvar{rstr}}{\desugar{\concrx{\digit+}}}$ that did not maintain the invariant that the string in the translation was in the language of the regular expression specified in the type index.
-Luckily, the representation of this type was held abstract from the perspective of the opcon structure of $\tcvar{lprod}$, so the translation, which would have otherwise been representationally consistent, did not pass the abstract representation check, because it was not \emph{representationally independent}, as enforced by our use of an internal type variable to stand for the representation.
-
-\begin{theorem}[Conservativity]If $\vdash \Phi \sim \kXi$ and $\stypeX{\sty{\tc}{\sttyidx}}$ and we have a value invariant ($\forall e$, if $\eana{\emptyset}{\Phi}{e}{\sty{\tc}{\sttyidx}}{\iota}$ then $\iota \mapsto^{*} \iota'$ and $\iota'~\mathtt{val}$ implies $P(\iota')$) and $\vdash \Phi, \phi \sim \kXi, \tc'[\kappa']$ then ($\forall e$, if $\eana{\emptyset}{\Phi, \phi}{e}{\sty{\tc}{\sttyidx}}{\iota}$ then $\iota \mapsto^{*} \iota'$ and $\iota'~\mathtt{val}$ implies $P(\iota')$).\end{theorem}
-
-Note that this is a stronger statement than stability, as $e$ may not have been well-typed before the extension (e.g.
-the operation $\opname{!ri}$).
-Our approach will be to introduce a technical device into the language that allows us to show, however, that there is always some other term with the same type and translation in the original closed world, precisely because  type constructors can only treat one another abstractly.
-Adding this technical device only weakens the set of properties that can be established benignly, in that one can no longer use exhaustiveness arguments to prove that the default branch of the $\keyw{tycase}$ construct is never taken inside an opcon definition.
-
-\begin{lemma}[Other]
-If $\vdash \Phi \sim \kXi$ and $\stypeX{\sty{\tc}{\sttyidx}}$ and $\eana{\Upsilon}{\Phi, \phi}{e}{\sty{\tc}{\sttyidx}}{\iota}$ then there exists an $e_\text{other}$ such that $\eana{\Upsilon}{\Phi}{e}{\sty{\tc}{\sttyidx}}{\iota}$.
-\end{lemma}
-\begin{proof}We proceed by induction on the typing derivation in the extended context.
-Most cases proceed by induction, so we omit the details.
-The interesting cases are for literals and targeted operations.
-
-For $e=\elit{\stmidx}{\es}$, we construct $e_\text{other}=\elit{\stmidx'}{\es'}$:
-\begin{itemize}
-\item We construct $\stmidx'$ from the value of $\stmidx$ by replacing every sub-term of the form $\sty{\tc'}{\stmidx''}$, where $\tc'$ is the new tycon defined by $\phi$, with $\sotherty{m}{\tau}$, where $\vdash_{\Phi, \phi} \sty{\tc'}{\stmidx''} \leadsto \tau$ and choosing an $m$ uniquely determined by $\stmidx''$ (needed below).
-The kinding rules then imply that $\sofkz{\emptyset}{\emptyset}{\kXi}{\stmidx'}{\kappa_\text{tmidx}}$, where $\kappa_\text{tmidx}$ is the literal index kind of $\tc$, because no other kinding rules for values depend on $\kXi$.
-This is why opcon indices were restricted to equality kinds.
-\item We construct $\es'$ based on the final argument store, $\memG$.
-In particular, for every entry in $\memG$ of the form $n \hookrightarrow e : \sty{\tc'}{\stmidx''} \leadsto \iota/x : \alpha$, we append an argument of the form  $\easc{\eother{\iota}}{\sotherty{m}{\tau}}$ to $\es'$, where $m$ is determined by the same mapping as above and $\tau$ is determined via the substitution induced by $\memD$.
-
-The kinding and normalization rules for $\sotherty{m}{\tau}$ imply that its representation is always held abstract, and the $\keyw{tycase}$ form always takes the ``else'' branch.
-This is precisely the same behavior that $\sty{\tc'}{\sttyidx}$ has, because $\tc' \neq \tc$ and the opcon definition was not kinded knowing of $\tc'$.
-This implies that, when starting with $\mkargs{\es'}{\stx{args}'}{\memG_0'}{n}$, the derivation of $e_\text{other}=\elit{\stmidx'}{\es'}$ will follow the same path, producing the same abstract translation, and the final argument store will differ only at the entries we modified: the $n$th entry will necessarily be of the form $n \hookrightarrow {\eother{\iota}} : {\sotherty{m}{\tau}} \leadsto \iota/x : \alpha$ where $\alpha$ maps to $\sotherty{m}{\tau}$ in the abstract rep store.
-As a result, the final substitution $\gamma$ maps every $x$ to its corresponding $\iota$, as in the derivation using extended context (in essence, we made the same translation arise from a different type for arguments that had a type constructed by the new tycon).
-\end{itemize}
-The argument for targeted operations follows an analagous route.
-Though a bit circuitous in detail, the essence of the proof is simple: because opcon definitions can only extract indices from  tycons that were already in $\Phi$, and because the representations and translations are held abstract by the semantics, we can simply introduce a generic family of ``other types'' and corresponding constants, each specifying its translation directly.
-These are distinct from the known types, so they cannot violate their invariants, but permit reasoning generically about future extensions.
-\end{proof}
-%If ** e <= ty[tc](styidx) ~> i then exists e_other s.t. e <= ty[tc](styidx)
-
-%onstruct a proof that $\eana{\emptyset}{\Phi}{e}{\sty{\tc}{\sttyidx}}{\iota}$ for every proof that $\eana{\emptyset}{\Phi, \phi}{e}{\sty{\tc}{\sttyidx}}{\iota}$.
-
-%These generally rely on monolithic specifications, so that exhaustive induction can be used in proofs.
-Combining specifications is a manual process.
-Chlipala and others have suggested that if these proofs are constructed by powerful automated procedures, then the burden of re-proving the metatheory may be decreased \cite{chlipala}.
-While we do not disagree, this approach comes with no guarantee that the automation procedure will succeed, only anecdotal evidence that the procedures are robust.
-We aim to render this ``hack'' unnecessary by making it possible to combine  proof terms derived separately directly and automatically.
-We also aim to provide such guarantees to  implementations, not just their specifications.
-
-\section{Related Work and Discussion}\label{prior-work}
-
-% Note also that type constructors are not themselves static values.
-To apply them in a curried manner or pass them around in the static language, we can define a static function, e.g.
-$\slam{\kty}{\svar{t1}}{\slam{\kty}{\svar{t2}}{\sparr{(\svar{t1}, \svar{t2})}}}$ has kind $\karrow{\kty}{\karrow{\kty}{\kty}}$.
-
-%To situate our work, and because we build on some recent results, we begin by summarizing prior approaches that can be considered when a simple isomorphic embedding of a fragment is not evident.
-
-\paragraph{Macros}
-If an isomorphic embedding can be established but it is too complex, language-integrated static term rewriting  (``macro'') systems, like Template Haskell \cite{SheardPeytonJones:Haskell-02} and Scala's static macros \cite{ScalaMacros2013}, can help by making it possible to generate ``boilerplate code'' automatically, moving the complexity of computing the embedding of a term into a metaprogram, but the static semantics are fixed.
-If each metaprogram is invoked like a function, inner macros are expanded before outer macros, and the mechanism enforces abstraction barriers by ensuring that the rewriting logic is hygienic, cannot modify surrounding code, and does not depend on shared state, then compositional reasoning is possible: the meaning of a term depends only on its subterms, not on the specific position it appears in a program.% Complex macros can, however, still be difficult to reason about, so many static macro systems impose further constraints (e.g.
-Scala recommends using only its ``black box'' macros, which  enforce a function-like typing discipline).
-These do not permit extensions to the underlying type system.
-%Rewriting systems that permit global pattern-based dispatch, however, do not admit strong modular reasoning principles (the same term might match multiple patterns, defined separately, creating ambiguities).
-
-\paragraph{Desugaring}\label{desugaring}
-Sometimes, dialects address issues of complexity by introducing new syntax.
-Indeed, most dialects do build in syntax for a few privileged abstractions (e.g.
-list literals are nearly ubiquitous, monad comprehensions support a key feature of Haskell).
-%For example, even though lists can be defined using datatypes, most languages build in special ``literal forms'' that make introducing and pattern matching over lists less tedious.
-Different dialects may similarly choose to  provide concise forms for working with other types of data (e.g.
-option types, regular expressions or SQL queries).
-Types not in the language's standard library, however, must typically use a more uniform syntax.
-%Evidence suggests that this leads programmers to choose less semantically meaningful representations (e.g. representing SQL queries as strings), which can cause a number of issues \cite{TSLs}. 
-To make the situation less asymmetric, systems that introduce new  ``desugarings'' atop a base language have been developed.
-For example, %Camlp4 is a language-external system used by the OCaml community.
-dialects of languages built using Sugar* \cite{erdweg2013framework} allow syntax extensions to be packaged separately and combined using language-integrated declarations.
-If the desugaring logic obeys the same constraints described above for macros, semantic reasoning can be performed compositionally (and recent work has shown how reasoning about type correctness can be automated \cite{conf/icfp/LorenzenE13}).
-
-These systems do not guarantee that the composition of unambiguous grammars will remain unambiguous (and realistic examples of ambiguous combinations are not uncommon, e.g.
-XML and HTML).
-%Every combination of desugarings is thus a  dialect with respect to syntactic reasoning.
-Recent work has made progress in addressing this problem by restricting how syntax extensions can interface with the host syntax.
-Schwerdfeger and Van Wyk require a globally unique start token and describe checkable conditions pertaining to the follow sets of host language non-terminals to guarantee that extensions can be composed unambiguously \cite{conf/pldi/SchwerdfegerW09}.
-This suffices for modularly adding new keyword-prefixed forms, but literal forms are awkward to define in this way.
-Omar et al.
-describe a language-integrated technique specifically for this purpose, using a technique with parallels to the one we are building on \cite{TSLs}.
-Literal parsing logic is directly associated with user-defined types, forming \emph{type-specific languages} (TSLs), and local type inference controls invocation of TSL parsing logic, guaranteeing that composition is unambiguous.
-The mechanism guarantees hygiene and inner terms cannot be inspected, so  fully modular reasoning is possible.
-
-%We assumed these desugaring techniques were available to solve issues strictly related to complexity and convenience.
-
-\paragraph{Optimization and Extensible Compilers}
-When an embedding that preserves the static semantics and is sufficiently simple exists, but a different embedding would better preserve a desired cost semantics, term rewriting techniques can also be used to perform specialized ``optimizations'', thus achieving an isomorphic embedding.
-Care must be taken, however, to ensure that the optimized value is not manipulated directly if the rewriting does not preserve the static semantics (i.e.
-the rewritten term has a different, less constrained type).
-Type abstraction can be used directly to achieve this property.
-\emph{Lightweight modular staging}, for example, uses Scala's support for abstract type members in traits to hide the optimized representations from the program \cite{Rompf:2012:LMS}.
-As long as optimizations are modularly known to be meaning-preserving (to be sure, a non-trivial proof obligation that benefits from mechanization, as in recent work on extensible compiler optimization \cite{conf/pldi/TatlockL10}), they can be composed arbitrarily.
-%As long as transformations are meaning-preserving (which can also be shown mechanically), they can be applied in any order.
-%The metaprograms can modularly be shown not to violate the type system of the host language, Modular reasoning is possible as long as dispatch is explicit (e.g.
-macros) or based on an unambiguous language mechanism (e.g.
-trait composition in LMS) and made simpler by the fact that transformations must be type-preserving.
-
-\paragraph{Type Refinement}
-%When T, the solution is less clear. 
-The techniques above require that an embedding that preserves the static semantics of a desired fragment already exists.
-When new static distinctions need to be made for values of an existing type, but new primitive operations are not needed, one solution is to develop a system of \emph{type refinements} based on the fragment of interest \cite{Freeman91}.
-For example, one might refine the type of integers to distinguish negative integers.
-Most proposals for ``pluggable type systems'' describe such type refinement systems \cite{Brac04a}, which require only additional annotations supplied as comments or metadata (e.g.
-JavaCOP \cite{Andreae:2006:FIP:1167473.1167479}).
-%Because the dynamic semantics are fixed,  composition is safe.
-Although our regular string types could perhaps be approximated with refinements, this does not permit the advanced operations, optimizations and representation independence results we enable.
-
-\paragraph{Language Frameworks}
-The most general situation is when exposing a fragment requires defining new types as well as new operators, where the static and dynamic semantics governing these new operators make non-trivial use of  statically valued information.
-We saw labeled product types added, which require a new type constructor and new operator constructors.
-The regular string type's group projection operator had a specialized cost semantics, so a simple refinement would not be able to approximate it.
-Scala's type system does not track the necessary invariants, so LMS-like optimizations would also not be appropriate.%For example, projection operators,  written concretely in some dialects of ML as \verb|#label|, can be seen as applications of \verb|#|, the projection operator constructor, to \verb|label|, a static index.
-The type of an operation like \verb|#label e| (sometimes written postfix as \verb|e#label|) is a function of the index of the type of \verb|e|.
-The dynamics also needs (some trace of) this information to determine how the operation is evaluated.
-
-%Many of the other examples mentioned above as motivations for new dialects have a similar flavor.
-We will detail another example in the next section of a type constructor that tracks strings known statically to be in a regular language, using it to dispense with unnecessary run-time checks in cases where they are not strictly necessary (affecting the cost component of the dynamic semantics).
-
-%An important observation is that while it is possible to \emph{implement} records by translation to tuples, or nested binary pairs, or lists, or many other types arising from a language like $\mathcal{L}\{{\rightharpoonup}\,{\forall}\,{\mu}\,{\keyw{1}}\,{\times}\,{+}\}$, it is not possible to isomorphically embed them as such (and thus rely on purely syntactic sugar) because there is no unique inverse mapping from, e.g., unlabeled product types to the corresponding record type (many combinations of labeled products may be implemented using the same unlabeled product).
-This is a common pattern even in more complex situations: a simple internal language (IL) generally suffices as a target for implementing the dynamics of an external language (EL) with a richer static semantics, but an embedding is not possible.
-This observation underlies the design of most compilers.
-
- % like SML, Specifying languages in a manner analagous to how they are usually implemented can simplify reasoning about the core of the language.
-
-% and typed translation semantics (the internal terms have a different type system).% Translation semantics
-
-%These types do not admit an isomorphic embedding using only unlabeled tuples because the field projection family of operators, indexed by labels (field projection is written concretely as \verb|#label(e)|, \verb|e#label| or \verb|e.label| in different languages), must be capable of operating on all values of such types.
-% with the labels removed would not admit a unique inverse.
-%(each equivalence class of record types could be implemented as an abstract type using tuples at considerable inconvenience but there would still be no way to define a projection operator that operated uniformly over all records having a particular field).
-
-% (i.e.
-fragments defining families of operations that cannot simply be written as a finite collection of functions, examples of which we will discuss).
-If there are a finite number of operations, and they can be typechecked without inspecting the structure of the types of the arguments, then an abstract type can be used.
-new primitive operations (including new variants of existing operations) that require static knowledge of these invariants, however, this too is infeasible.
-take advantage of the invariants being maintained, and these operations must be usable throughout a program, .
-%At best, annotations might be written as ``comments'' or metadata for use by these tools.
-
-A variety of \emph{language frameworks} have been developed to make it easier to define new languages, in some cases together with a compiler and other tooling for these languages.
-Various terms have been used in the literature, including \emph{compiler generators} and \emph{language workbenches}; see \cite{erdweg2013state}.
-%Some \emph{extensible compilers} are also actually language frameworks, because they permit the introduction of new constructs, not just new meaning-preserving optimizations.
-The maintainers of many compilers, e.g.
-Haskell's GHC, treat their compiler as a ``laboratory'' for new fragments, toggled by flags or special comments.
-\emph{Logical frameworks} focus more specifically on helping language designers mechanize the metatheory of language and logic specifications.
-T
-hese tools can decrease the effort needed to define a new language, but they either do not support forming languages from separately defined fragments or provide few modular reasoning principles that make it possible to reason separately about these fragments.
-Every combination of fragments is a new dialect, and must be reasoned about monolithically.
-%Efforts to reason modularly areeed, several problems can come up when fragments are combined, so .
-Let us briefly review difficulties that arise.
-
-%Concrete syntax known to be separately unambiguous might not be unambiguous when combined, as was already discussed above. 
-If the abstract syntax needs to be extended to support a new fragment, problems also arise.
-In monolithic settings, terms can be implemented using finite recursive sums (i.e.
-term constructors are often implemented as ML-style datatype constructors), but this does not permit extension, so open sum types or products of functions (i.e.
-objects \cite{conf/oopsla/Aldrich13}) must instead be used.
-This can present issues when one wishes to modularly define new functionality that should exhaustively cover all terms in the language (e.g.
-pretty-printers for expressions).
-Reynolds first identified this problem \cite{Reynolds75} and Wadler named it the \emph{expression problem}.
-%A number of language frameworks (e.g.
-JastAdd \cite{Ekman:2007:JEJ:1297027.1297029} and Silver \cite{VanWyk:2010:SEA}) use extensible \emph{attribute grammars}.
-These can be seen as a form of open sum, where attributes correspond to functions performing traversals (more specifically, \emph{catamorphisms} \cite{catamorphisms}\todo{citation / remove?}).
-%When a new term constructor is added, the logic determining how some existing attributes (e.g.
-typechecking and translation) should handle the new case is also provided.
-\emph{Forwarding} can be used to attempt to delegate responsibility over attributes other than those explicitly defined to another term constructor, eventually leading to one in the fixed internal language \cite{VanWyk:2010:SEA}.
-This can address some aspects of the expression problem, but creates a bigger problem: the internal implementation details  of a fragment are necessarily exposed, violating an abstraction barrier that, as we will discuss, is critical.
-In this work, we sidestep the problems of syntax, instead leaving it fixed and relying on a bidirectional type system delegating to a relevant tycon.
-%For example, projection operators do not require adding a corresponding term constructor to the abstract syntax.
-Instead, projection is categorized as a \emph{targeted operation}, so  the concrete term \verb|#label e| desugars to an abstract term like $\keyw{targ}[(\desugar{\conclbl{\#}}, \desugar{\conclbl{\texttt{label}}})](e)$ (where $\desugar{\cdot}$ denotes an encoding of constant labels, here a label corresponding to the operator constructor and the field label itself, into the static language).
-The type constructor of the type synthesized by the target argument, here $e$, determines the semantics of the term.
-If a new record-like fragment is added (e.g.
-they give the example of one with prototypic inheritance), it can reuse the same concrete and abstract syntax directly, precluding conflicts and avoiding the need to define the behavior of tools like pretty printers for every new fragment.
-As in their work on type-specific languages, described above, the semantics of literal forms are controlled by the type constructor of the type the literal is being analyzed against.
-%We will discuss this in the next section.
-
-Once syntactic issues have been addressed, however, there are a host of semantic guarantees  that must be established before a language can be relied upon, as we saw.
-% The most basic guarantee is \emph{type safety}: that the dynamics are well-defined for all terms accepted as well-typed by the statics, and preserve the statics during evaluation.
-Each fragment can then build on type safety to establish additional \emph{type invariants} stating properties about all types that it constructs, which  clients can then use to reason about programs (e.g.
-value induction for eager sum types relies on finiteness of the canonical forms).
-In practice, typechecking is expected be \emph{deterministic} and \emph{decidable} (i.e.
-terminating) and, except in circumstances where non-determinism is explicitly exposed to programmers, the dynamic semantics are also expected to be deterministic.
-%Precise formulations of these properties depend on how the semantics are specified.%, so we will make this more precise below.
-Modern language frameworks guarantee few or none of these properties about the dialects they produce.
-More alarmingly, even when these properties have been established for two dialects  (either in the metatheory or mechanically using a logical framework), and syntactic conflicts are addressed, there is no guarantee that merging the dialects together will conserve these properties.
-We saw examples of such failures of modularity above.
-If our goal is to integrate fragment composition into the language, these tools are thus of limited utility.
-Clients would have to take on the burden of reasoning about these basic properties for every combination of ``libraries'' they chose to import.% Improving this state of affairs, so that it more resembles reasoning about separately defined modules (we assume an ML-style module system), is the general topic of this paper.
-
-\begin{figure*}[t]
-\small
-\newcommand{\twocol}[1]{\multicolumn{2}{l}{#1}}
-\newcommand{\threecol}[1]{\multicolumn{3}{l}{#1}}
-\hspace{-5px}$\begin{array}{llll}
-\svar{arity0} : \klist{\karg} \rightarrow \kunit & \svar{str\_of\_Str} : \kstr \rightarrow \kitm & \svar{uniqmap} : \klist{\klbl \times \kty}\rightarrow \kunit &  \svar{lbleq} : \klbl \rightarrow \klbl \rightarrow \kunit\\
-\svar{arity1} : \klist{\karg} \rightarrow \karg & \svar{rstrs\_of\_Strs} : \klist{\kstr} \rightarrow \kitm & {\svar{lookup} : \klist{\klbl\times\kty}\rightarrow \klbl \rightarrow (\knat \times \kty)}\\
-\svar{arity2} : \klist{\karg} \rightarrow \karg \times \karg & \svar{rx\_of\_Rx} : \krx \rightarrow \kitm & \svar{rmatch} : \krx \rightarrow \kstr \rightarrow (\kstr \times \klist{\kstr})\\
-\svar{nil} : \kforall{\kalpha}{\klist{\kalpha}} & \svar{nat\_of\_Nat} : \knat \rightarrow \kitm & \svar{rconcat} : \krx \rightarrow \krx \rightarrow \krx & \hspace{-15px}rsconcat : \iparr{\iparr{\keyw{rs}}{\keyw{rs}}}{\keyw{rs}}\\
-\svar{list1} : \kforall{\kalpha}{\kalpha \rightarrow \klist{\kalpha}} & \svar{prjnth} : \knat \rightarrow \kitm \rightarrow \kitm & \svar{rgroupn} : \knat \rightarrow \krx \rightarrow \krx &\hspace{-15px} rsgroupn : \iparr{\keyw{nat}}{\iparr{\keyw{rs}}{\keyw{rs}}}\\
-\svar{list2} : \kforall{\kalpha}{\kalpha \rightarrow \kalpha \rightarrow \klist{\kalpha}} & \svar{append} : \kforall{\kalpha}{\klist{\kalpha}\rightarrow \kalpha \rightarrow \klist{\kalpha}} & \svar{rsublang} : \krx \rightarrow \krx \rightarrow \kunit & \hspace{-15px}rsregroup : \iparr{\keyw{rx}}{\iparr{\keyw{rs}}{\keyw{rs}}}\\
-\twocol{\svar{listrec} : \kforall{\kalpha_1}{\kforall{\kalpha}{\klist{\kalpha_1}\rightarrow \kalpha \rightarrow (\kalpha_1\rightarrow\kalpha\rightarrow\kalpha)\rightarrow\kalpha}}} & rscheck : \iforall{t}{\iparr{\keyw{rx}}{\iparr{\keyw{rs}}{\iparr{(\iparr{\keyw{rs}}{t})}{t}}}}\\
-\threecol{\svar{listrec2} : \kforall{\kalpha_1}{\kforall{\kalpha_2}{\kforall{\kalpha}{\klist{\kalpha_1}\rightarrow\klist{\kalpha_2}\rightarrow \kalpha \rightarrow (\kalpha_1\rightarrow\kalpha_2\rightarrow\kalpha\rightarrow\kalpha)\rightarrow\kalpha}}}}\\
-\threecol{\svar{listrec3} : \kforall{\kalpha_1}{\kforall{\kalpha_2}{\kforall{\kalpha_3}{\kforall{\kalpha}{\klist{\kalpha_1}\rightarrow\klist{\kalpha_2}\rightarrow \klist{\kalpha_3}\rightarrow\kalpha \rightarrow (\kalpha_1\rightarrow\kalpha_2\rightarrow\kalpha_3\rightarrow\kalpha\rightarrow\kalpha)\rightarrow\kalpha}}}}}\\
-\threecol{\svar{syn} := \scolorc\lambda \svar{a}{:}\karg.\keyw{letpair}~(\svar{synfn}, \_) = \svar{a}~\keyw{in}~\svar{synfn}~\striv}\\
-\threecol{\svar{ana} := \scolorc\lambda \svar{a}{:}\karg.\keyw{letpair}~(\_, \svar{anafn})=\svar{a}~\keyw{in}~\svar{anafn}}
-\end{array}$
-\caption{Definitions and signatures of helpers used in the examples.}
-\label{helper-sigs}
-\end{figure*}
-
-Though we have not mechanically proven these properties for our design, we have given strong evidence that these properties are maintained modularly.
-That is, we need only establish the semantics of the opcon structures in isolation.
-%We can then rely on, e.g., the Conservativity theorem above.
-The most immediate direction for future work is to finishing mechanizing the metatheory for this  technique, and to embed it into an existing proof assistant (using a continuation passing style to simulate the stores in our normalization semantics).
-Despite this, we believe that the technique as described semi-formally above is compelling.
-Adding support for tycon-specific typing contexts, implicit coercions and direct opcon calls can all be taken up by the community.
-Delving further into the question of when can two tycons with the same signature be substituted for one another, using a technique based on admissible relations, is also an avenue we wish to explore \cite{pfpl}.
-
-%We note that is increasingly being deployed in contemporary languages (e.g.
-Scala) due to the perception that it enables good error messages, and to support semantics where  non-local type inference is undecidable.
-Permitting extensibility together with non-local type inference is an open problem.
-
-Our work has some similarities to work on maintaining type abstraction using a form of effect system \cite{CraryHarperDreyer2002}, and on translating modules to System F \cite{conf/tldi/RossbergRD10}.
-Unlike modules, tycons permit the representation types to be computed, use a separate IL, and permit operations whose types cannot simply be written as functions.
-%Both examples we used here would be quite difficult to fully embed using only modules.
-
-%\section{Conclusion}\label{conclusion}
-
-%\section{Older Stuff}
-%
-%When only a weak embedding is possible, the solution is less clear. 
-%
-%
-%
-%
-%even in the rare situation where a common framework has been used or the dialects are implemented within the same compiler, as these mechanisms do not guarantee that combining separately sound dialects will be sound.
-Avoiding dialect composition is difficult, because interactions between components of an application written in different dialects leads to precisely the situation in item 3 above.% We will consider this further in our discussion of related work in Sec.
-\ref{related-work}.
-%
-%\paragraph{Language-Integrated Approaches}
-%
-%\paragraph{Active Type Constructors}
-%
-%\paragraph{Contributions}
-%
-% For example, the  fragment providing binary product types might define forms like $\texttt{(\textit{e}}\texttt{, \textit{e}}\texttt{)}$, $\texttt{$e$}\texttt{.l}$ and $\texttt{$e$}\texttt{.r}$, the term constructors $\FF{pair}$ and $\FF{prj}$ and the type constructor $\fvar{prod}$.
-A \emph{term} is constructed by applying a term constructor to an \emph{index} and zero or more \emph{arguments}.
-For example, \emph{expressions}, $e$, which are one \emph{sort} of term, are written $\FF{pair}[\FF{triv}](e_1; e_2)$, $\FF{prj}[\FF{left}](e)$ and $\FF{prj}[\FF{right}](e)$.
-Expressions are indexed by \emph{static terms}, $\sigma$.
-These are often simply assumed to be part of the ambient metatheory, but we will specify static term constructors explicitly.
-We omit trivial indices and empty argument lists for terms other than expressions for concision.
-%
-%Types arise by applying a type constructor to a \emph{type index}, also a static term.
-For example, the type constructor $\fvar{prod}$ is indexed by a static pair of types (for $n$-ary products, the index would instead be a static list of types).
-The sort where types are defined is commonly referred to as the \emph{type-level language}, but because types can appear in term and type indices, it is useful for us to include them directly in the static language.
-We might specify a trivially indexed static term constructor corresponding to each type constructor where the type index is the term argument, writing $\FF{prod}(\FF{spair}(\sigma_1; \sigma_2))$, or we might define a single static term constructor $\FF{ty}$ indexed by a type constructor, writing $\FF{ty}[\fvar{prod}](\FF{spair}(\sigma_1; \sigma_2))$.
-The static and dynamic semantics can be specified in a standard way \cite{pfpl}.
-%The static semantics serve as a specification for the typechecker and the dynamic semantics serves as a specification for the first stage of a compiler, which we call translation.
-In some cases, the static and dynamic semantics are given simultaneously by giving translation to an internal language (e.g.
-the Harper-Stone semantics of Standard ML \cite{harper-stone}).
-% Note that in practice, the dynamic semantics of expressions is implemented by translation to a simpler \emph{internal language}, with terms $\iota$ and types $\tau$.
-%
-%
-%
-%%Modern languages like ML provide general-purpose type constructors that admit strong and often satisfying embeddings of many language fragments, occupying what is widely seen as a ``sweet spot'' in the design space.
-%
-%New fragments are generally introduced as distinct \emph{dialects} of an existing general-purpose language, constructed either as a fork, using tools like compiler generators, DSL frame\-works or  language workbenches, or directly within a particular compiler for the language, often activated by a flag or pragma.
-%These are not the sorts of problems usually faced by library providers.
-Well-designed languages preclude the  possibility of ``link-time'' conflicts between libraries and ensure that the semantics of one library  cannot be weakened by another by strictly enforcing abstraction barriers.
-For example, a module declaring an abstract type in ML can rely on its representation invariants no matter which other modules are in use, so clients can assume that they will operate robustly in combination without needing to attend to burdensome ``link-time'' proof obligations.
-%
-%
-%
-%
-% number of researchers have observed that the problem of combining fragments to form a language is made complicated by problems of \emph{ambiguity}.
-The simplest forms of ambiguities are naming ambiguities: two type constructors or term constructors may have the same name.
-Globally unique name generation mechanisms are known to avoid this problem.
-%
-%
-%Types arise by applying a type constructor to a \emph{type index}.
-We will view both types and type indices as  static terms.
-Types arise via the static term constructor $\FF{ty}$, which is indexed by a type constructor and has a  single argument, the type index, e.g.
-$\FF{ty}[\fvar{prod}](\FF{spair}(\sigma_1; \sigma_2))$, where $\sigma_1$ and $\sigma_2$ are two other types and $\FF{spair}$ is the static term constructor for static pairs (its trivial index is omitted).
-In the concrete syntax, we might write this instead $\fvar{prod}\texttt{[}\sigma_1; \sigma_2\texttt{]}$ or using a stylized infix notation, $\sigma_1 \times \sigma_2$.
-%
-%A fragment is often identified by the type constructor it is organized around, so the STLC is  called $\mathcal{L}\{\fvar{arrow}\}$, following the convention used by Harper \cite{pfpl} and others (but avoiding ``stylistic sugar'' for the sake of uniformity).
-%
-%G\"odel's $\mathbf{T}$ consists of the $\rightarrow$ fragment and the \verb|nat| fragment, defining natural numbers and a recursor that allows one to ``fold'' over a natural number.
-We might thus call it $\mathcal{L}\{\rightarrow \mathtt{nat}\}$.
-This language is  more powerful than the STLC because the STLC admits an embedding  into  $\mathbf{T}$ but the reverse is not true.
-Buoyed by this fact, we might go on by adding fragments that define sums, products and various forms of inductive types (e.g.
-lists), or perhaps a general mechanism for defining inductive types.
-Each fragment clearly increases the expressiveness of our language.
-%
-%
-%If we consider the $\forall$ fragment, however, defining universal quantification over types (i.e.
-parametric polymorphism), we must take a moment to reflect.
-In $\mathcal{L}\{\rightarrow\,\forall\}$, studied variously by Girard as System $\mathbf{F}$ \cite{girard1971extension} and Reynolds as the polymorphic lambda calculus \cite{Reynolds94anintroduction}, it is known that sums, products, and inductive and co-inductive types can all be weakly defined.
-This means that we can \emph{translate} well-typed terms of a language like $\mathcal{L}\{\rightarrow\forall~\mathtt{nat} + \times\}$ to well-typed terms in $\mathcal{L}\{\rightarrow\forall\}$ in a manner that preserves their dynamic semantics.
-But Reynolds, in a remark that recalls the ``Turing tarpit'' of Perlis \cite{Perl82a}, reminds us that programming with the corresponding \emph{embedding} may be unwise \cite{Reynolds94anintroduction}:
-%\begin{quote}
-%To say that any reasonable function can be expressed by some program is not to say that it can be expressed by the most reasonable program.
-It is clear that the language requires a novel programming style.
-Moreover, it is likely that certain important functions cannot be expressed by their most efficient algorithms.
-%\end{quote}
-%
-%%Adding new fragments directly to a language can make statically reasoning about programs more precise, support a more natural programming style and endow the language with a more favorable cost semantics.
-The latter two points might be relevant even if a strong embedding (i.e.
-one where there is a semantics-preserving isomorphism between terms and types of the fragment and the language) can be found.
-%%
-%%Consistent with this view, typed programming languages like ML expose, for example, both $n$-ary product types and record types, despite the fact that any product type is definable using a record type (or in the other direction, using a module to provide the field selection operator).
-The compiler, on the other hand, \emph{is} mainly concerned with implementing the dynamic semantics correctly, and indeed, compilers often use the fact that weak definability results are quite readily derivable to their advantage, translating the constructs of an external language (EL) to those of a much simpler internal language (IL), e.g.
-with only binary products, during or directly after typechecking\todo{cite Harper-Stone}.
-%Correctly implementing the dynamic semantics of the language becomes the primary concern during this phase, so a simple IL that decreases the number of cases needing consideration when implementing and verifying the compiler is a wise choice.
-%
-%%Having established why a minimal language like $\mathcal{L}\{\rightarrow\forall\}$, while suitable as an internal language, needs to be extended with additional fragments before it is suitable for use as a human-facing external language,
-%A stronger embedding requires constructing an isomorphism that preserves both static reasoning principles and dynamic behaviors, in a sense that we will make more precise as we go on.
-This is generally far more difficult to establish than a weak embedding, and fewer universality results can be applied.
-For example, the aforementioned embeddings into $\mathcal{L}\{\rightarrow\forall\}$ do not preserve type disequality and some other equational reasoning principles (we will see less subtle issues soon).
-Establishing an isomorphic embedding that is natural (as measured, to a first approximation, by the ratio of boilerplate code that must be manually generated in the embedding vs.
-in the fragment, and how likely one is to make a mistake when writing it) and that has a reasonable cost semantics is harder still.
-%
-%
-%\paragraph{Contributions}
-%Drawing from this approach, we aim to introduce a mechanism that makes it possible to declare new type and operator constructors and define, \emph{using type-level functions}, their static and dynamic semantics, the latter by translation to a fixed typed internal language.
-If a language fragment can be weakly defined in terms of the typed internal language and its statics are of a general form consistent with this protocol, a n isomorphic embedding in the external language can be constructed by simply introducing the necessary constructors and defining directly the necessary type assignment and translation logic.
-We call such an embedding an \emph{active embedding}.
-%
-%The semantics imposes various checks to ensure type safety and decidability of type assignment and equality.
-Moreover, an important class of lemmas that play a key role in  proofs that a library enables strong embeddings of a language fragment can be derived in a suitable ``closed world'' (that is, in the traditional manner where one reasons inductively over a \emph{finite} collection of constructors).
-The semantics guarantees that they will be \emph{conserved} in the  ``open world'' by enforcing an abstraction barrier at extension boundaries.
-This avoids the most fundamental semantic problems that can arise when composing dialects and justifies the inclusion of the mechanism inside the language.
-%
-%We will begin in Sec.
-\ref{core} by introducing a minimal calculus with this kind of {actively typed} semantics, @$\lambda$.
-The external language begins with only the $\rightarrow$ fragment and the typed internal language is a simple variant of PCF that permits only weak embeddings of even simple fragments like the $\mathtt{nat}$ fragment of G\"odel's $\mathbf{T}$ or ML-style $n$-ary products.
-We actively embed these into the external language to explain the mechanism and motivate the constraints that the semantics imposes.
-We then examine the semantics in greater detail and state and sketch the key points in the proofs of the metatheory in Sec.
-\ref{theory}.
-These two sections comprise the fundamental contributions of this work.
-%
-%While the core calculus addresses the issue of making a strong embedding possible, active embeddings into the core calculus are often impractical syntactically.
-To begin to address the issue of syntax, we develop in Sec.
-\ref{expanded-syntax} a flexible type-directed dispatch protocol for a more conventional {expanded syntax} that defers to the core calculus semantically.
-We then use the expanded syntax to discuss several more interesting examples in Sec.
-\ref{examples}: labeled sums and products, the latter with delegation and functional field update, and a variant of \verb|sprintf|.\todo{update this if something doesn't make it}
-%
-%While of surprising expressiveness given its minimality, @$\lambda$ itself is certainly not capable of admitting embeddings of the full variety of type system fragments enumerated earlier.
-Indeed, to guarantee safety and conservativity and because our intention in this paper is to cleanly expose the essence of this mechanism, our calculus makes perhaps too many assumptions about the type system being embedded.
-For example, it only admits fragments where the typing judgement looks essentially like the typing judgement of the $\rightarrow$ fragment, so new contexts cannot be introduced.
-We discuss this and some of the other reasons why a fragment may not be satisfyingly embeddable in Sec.
-\ref{limitations}.
-These limitations do not appear fundamental, so we also outline directions for future work.
-We conclude by considering related work in Sec.
-\ref{related-work}.
-%
-%\section{Core @$\lambda$}\label{core}
-%%\newcommand{\pimport}[2]{\keyw{import}~#1~\keyw{in}~#2}
-%%
-%%\newcommand{\es}{\overline{e}}
-%%\newcommand{\elet}[3]{\keyw{let}(#1; #2.#3)}
-%%\newcommand{\eslet}[3]{\keyw{slet}[#1](#2.#3)}
-%%\newcommand{\easc}[2]{\keyw{asc}[#1](#2)}
-%%\newcommand{\elamrec}[3]{\keyw{lamrec}(#1,#2.#3)}
-%%\newcommand{\elit}[2]{\keyw{lit}[#1](#2)}
-%%\newcommand{\etarg}[3]{\keyw{targ}[#1](#2; #3)}
-%%\newcommand{\ebranch}[3]{\keyw{branch}[#1](#2; #3)}i
-%%
-%%\newcommand{\qitm}{\grave\iota}
-%%\newcommand{\qs}{\overline{\qitm}}
-%%\newcommand{\qlet}[3]{\keyw{let}(#1; #2.#3)}
-%%\newcommand{\qasc}[2]{\keyw{asc}[#1](#2)}
-%%\newcommand{\qlamrec}[3]{\keyw{lamrec}(#1,#2.#3)}
-%%\newcommand{\qlit}[2]{\keyw{lit}[#1](#2)}
-%%\newcommand{\qtarg}[3]{\keyw{targ}[#1](#2; #3)}
-%%\newcommand{\qap}[2]{\keyw{ap}(#1; #2)}
-%%\newcommand{\qstr}[1]{\keyw{str}[#1]}
-%%\newcommand{\qstrrec}[4]{\keyw{strrec}(#1; #2; {#3}.#4)}
-%%\newcommand{\qpair}[2]{\keyw{pair}(#1; #2)}
-%%\newcommand{\qletpair}[4]{\keyw{letpair}(#1; {#2}, {#3}.#4)}
-%%\newcommand{\qnil}{\keyw{nil}}
-%%\newcommand{\qcons}[2]{\keyw{cons}(#1; #2)}
-%%\newcommand{\qlistrec}[6]{\keyw{listrec}(#1; #2; {#3}, {#4}, {#5}.#6)}
-%%\newcommand{\qderivate}[3]{\keyw{derivate}[#1; #2; #3]}
-%%\newcommand{\quq}[1]{{\lhd}[#1]}
-%%
-%%\newcommand{\hitm}{\hat{\iota}}
-%%\newcommand{\hs}{\overline{\hitm}}
-%%\newcommand{\hlet}[3]{\keyw{let}(#1; #2.#3)}
-%%\newcommand{\hasc}[2]{\keyw{asc}[#1](#2)}
-%%\newcommand{\hlamrec}[3]{\keyw{lamrec}(#1,#2.#3)}
-%%\newcommand{\hlit}[2]{\keyw{lit}[#1](#2)}
-%%\newcommand{\htarg}[3]{\keyw{targ}[#1](#2; #3)}
-%%\newcommand{\hbranch}[3]{\keyw{branch}[#1](#2; #3)}
-%%\newcommand{\hap}[2]{\keyw{ap}(#1; #2)}
-%%\newcommand{\hstr}[1]{\keyw{str}[#1]}
-%%\newcommand{\hstrrec}[4]{\keyw{strrec}(#1; #2; {#3}.#4)}
-%%\newcommand{\htriv}{\keyw{triv}}
-%%\newcommand{\hpair}[2]{\keyw{pair}(#1; #2)}
-%%\newcommand{\hletpair}[4]{\keyw{letpair}(#1; {#2}, {#3}.#4)}
-%%\newcommand{\hinl}[2]{\keyw{inl}[#1](#2)}
-%%\newcommand{\hinr}[2]{\keyw{inr}[#1](#2)}
-%%\newcommand{\hcase}[5]{\keyw{case}(#1; #2.#3; #4.#5)}
-%%\newcommand{\hnil}{\keyw{nil}}
-%%\newcommand{\hcons}[2]{\keyw{cons}(#1; #2)}
-%%\newcommand{\hlistrec}[6]{\keyw{listrec}(#1; #2; {#3}, {#4}, {#5}.#6)}
-%%\newcommand{\hderivate}[3]{\keyw{derivate}[#1; #2; #3]}
-%%\newcommand{\huq}[1]{{\lhd}[#1]}
-%%
-%%\newcommand{\itm}{\iota}
-%%\newcommand{\is}{\overline{itm}}
-%%\newcommand{\ilet}[3]{\keyw{let}(#1; #2.#3)}
-%%\newcommand{\iasc}[2]{\keyw{asc}[#1](#2)}
-%%\newcommand{\ilamrec}[3]{\keyw{lamrec}(#1,#2.#3)}
-%%\newcommand{\iap}[2]{\keyw{ap}(#1; #2)}
-%%\newcommand{\istr}[1]{\keyw{str}[#1]}
-%%\newcommand{\istrrec}[4]{\keyw{strrec}(#1; #2; {#3}.#4)}
-%%\newcommand{\itriv}{\keyw{triv}}
-%%\newcommand{\ipair}[2]{\keyw{pair}(#1; #2)}
-%%\newcommand{\iletpair}[4]{\keyw{letpair}(#1; {#2}, {#3}.#4)}
-%%\newcommand{\iinl}[2]{\keyw{inl}[#1](#2)}
-%%\newcommand{\iinr}[2]{\keyw{inr}[#1](#2)}
-%%\newcommand{\icase}[5]{\keyw{case}(#1; #2.#3; #4.#5)}
-%%\newcommand{\inil}{\keyw{nil}}
-%%\newcommand{\icons}[2]{\keyw{cons}(#1; #2)}
-%%\newcommand{\ilistrec}[6]{\keyw{listrec}(#1; #2; {#3}, {#4}, {#5}.#6)}
-%%
-%%%\newcommand{\scolor}[1]{{\color{blue} #1}}
-%%\newcommand{\st}{\sigma}
-%%\newcommand{\stx}[1]{\st_\text{#1}}
-%%\newcommand{\sval}{\hat\sigma}
-%%\newcommand{\svalx}[1]{\sval_\text{#1}}
-%%\newcommand{\sx}[1]{{\bf #1}}
-%%\newcommand{\slam}[3]{\lambda[#1](\sx{#2}.#3)}
-%%\newcommand{\sap}[2]{\keyw{ap}(#1; #2)}
-%%\newcommand{\striv}{\keyw{triv}}
-%%\newcommand{\sinl}[2]{\keyw{inl}[#1](#2)}
-%%\newcommand{\sinr}[2]{\keyw{inr}[#1](#2)}
-%%\newcommand{\scase}[5]{\keyw{case}(#1; \sx{#2}.#3; \sx{#4}.#5)}
-%%\newcommand{\sfold}[3]{\keyw{fold}[#1.#2](#3)}
-%%\newcommand{\srec}[4]{\keyw{rec}[#1](#2; \sx{#3}.#4)}
-%%\newcommand{\sstr}[1]{\keyw{str}[#1]}
-%%\newcommand{\sstrrec}[4]{\keyw{strrec}(#1; #2; \sx{#3}.#4)}
-%%\newcommand{\spair}[2]{\keyw{pair}(#1; #2)}
-%%\newcommand{\sletpair}[4]{\keyw{letpair}(#1; \sx{#2}, \sx{#3}.#4)}
-%%\newcommand{\snil}{\keyw{nil}}
-%%\newcommand{\scons}[2]{\keyw{cons}(#1; #2)}
-%%\newcommand{\slistrec}[6]{\keyw{listrec}(#1; #2; \sx{#3}, \sx{#4}, \sx{#5}.#6)}
-%%\newcommand{\sty}[3]{\keyw{ty}[\fvar{#1}](#2.#3)}
-%%\newcommand{\selfvar}[1]{\textrm{#1}}
-%%\newcommand{\m}{\selfvar{m}}
-%%\newcommand{\stycase}[5]{\keyw{tycase}[\fvar{#1}](#2; \sx{#3}.#4; #5)}
-%%\newcommand{\sarg}[1]{\keyw{arg}[#1]}
-%%%\newcommand{\sanaarg}[5]{\keyw{ana}(#1; #2; \sx{#3},\sx{#4}.#5)}
-%%%\newcommand{\ssynarg}[3]{\keyw{syn}(#1; \sx{#2}.#3)}
-%%\newcommand{\sana}[3]{{\color{gray}\keyw{ana}[#1](#2; #3)}}
-%%\newcommand{\ssyn}[2]{{\color{gray}\keyw{syn}[#1](#2)}}
-%%\newcommand{\sraise}{\keyw{raise}}
-%%\newcommand{\strtm}[1]{{\rhd}(#1)}
-%%\newcommand{\strty}[1]{{\blacktriangleright}(#1)}
-%%\newcommand{\saitm}[1]{{\hat\rhd}(#1)}
-%%\newcommand{\saity}[1]{{\hat\blacktriangleright}(#1)}
-%%\newcommand{\styeq}[4]{\keyw{tyeq}(#1; #2; #3; #4)}
-%%
-%%\newcommand{\kx}[1]{\kappa_\text{#1}}
-%%\newcommand{\karrow}[2]{#1 \rightarrow #2}
-%%\newcommand{\kstr}{\keyw{Str}}
-%%\newcommand{\kprod}[2]{#1 \times #2}
-%%\newcommand{\ksum}[2]{#1 + #2}
-%%\newcommand{\kunit}{\keyw{1}}
-%%\newcommand{\klist}[1]{\keyw{List}[#1]}
-%%\newcommand{\kind}[2]{\keyw{ind}(#1.#2)}
-%%\newcommand{\kty}{\keyw{Ty}}
-%%\newcommand{\karg}{\keyw{Arg}}
-%%\newcommand{\ktrtm}{\keyw{TrTm}}
-%%\newcommand{\ktrty}{\keyw{TrTy}}
-%%
-%%\newcommand{\qity}{\grave\tau}
-%%\newcommand{\qtparr}[2]{#1 \rightharpoonup #2}
-%%\newcommand{\qtstr}{\keyw{str}}
-%%\newcommand{\qtprod}[2]{#1 \times #2}
-%%\newcommand{\qtlist}[1]{\keyw{list}[#1]}
-%%\newcommand{\qtrepof}[1]{\keyw{rep}[#1]}
-%%\newcommand{\qtuq}[1]{{\blacktriangleleft}[#1]}
-%%
-%%\newcommand{\hity}{\hat\tau}
-%%\newcommand{\htparr}[2]{#1 \rightharpoonup #2}
-%%\newcommand{\htstr}{\keyw{str}}
-%%\newcommand{\htunit}{\keyw{1}}
-%%\newcommand{\htprod}[2]{#1 \times #2}
-%%\newcommand{\htsum}[2]{#1 + #2}
-%%\newcommand{\htrec}[2]{\mu#1.#2}
-%%\newcommand{\htlist}[1]{\keyw{list}[#1]}
-%%\newcommand{\htrepof}[1]{\keyw{rep}[#1]}
-%%\newcommand{\htuq}[1]{{\blacktriangleleft}[#1]}
-%%
-%%\newcommand{\ity}{\tau}
-%%\newcommand{\itparr}[2]{#1 \rightharpoonup #2}
-%%\newcommand{\itunit}{\keyw{1}}
-%%\newcommand{\itstr}{\keyw{str}}
-%%\newcommand{\itsum}[2]{#1 + #2}
-%%\newcommand{\itrec}[2]{\mu#1.#2}
-%%\newcommand{\itprod}[2]{#1 \times #2}
-%%\newcommand{\itlist}[1]{\keyw{list}[#1]}
-%%
-%%\newcommand{\esyn}[5]{#1 \vdash_{#2} #3 \Rightarrow #4 \leadsto #5}
-%%\newcommand{\esynX}[3]{\esyn{\Upsilon}{\Phi}{#1}{#2}{#3}}
-%%
-%%\newcommand{\eana}[5]{#1 \vdash_{#2} #3 \Leftarrow #4 \leadsto #5}
-%%\newcommand{\eanaX}[3]{\eana{\Upsilon}{\Phi}{#1}{#2}{#3}}
-%%
-%%\newcommand{\etyerr}[3]{#1 \vdash_{#2} #3~\mathtt{tyerr}}
-%%\newcommand{\etyerrX}[1]{\etyerr{\Upsilon}{\Phi}{#1}}
-%%
-%%\newcommand{\efragerr}[3]{#1 \vdash_{#2} #3~\mathtt{ferr}}
-%%\newcommand{\efragerrX}[1]{\efragerr{\Upsilon}{\Phi}{#1}}
-%%
-%%\newcommand{\sevalCtx}{\mathcal{C}}
-%%\newcommand{\sevalCtxNone}{\cdot}
-%%\newcommand{\seval}[3]{#1 \Downarrow_{#2} #3}
-%%\newcommand{\sevalC}[2]{\seval{#1}{\sevalCtx}{#2}}
-%%\newcommand{\sevalN}[2]{\seval{#1}{\sevalCtxNone}{#2}}
-%%
-%%\newcommand{\tctxbind}[2]{#1 \Rightarrow #2}
-%%
-%%\newcommand{\kctx}{\Psi}
-%%\newcommand{\skind}[4]{#1 \vdash_{#2} #3 : #4}
-%%\newcommand{\skindX}[2]{\skind{\kctx}{\Phi}{#1}{#2}}
-%%\newcommand{\args}{\mathtt{args}(\es)=\svalx{args};\Xi}
-%%
-%%\begin{figure*}[t]
-%%\small
-%%\hspace{-5px}
-%%$
-%%\begin{array}[t]{rlcl}
-%%\mvlbl{programs} &
-%%p & ::= &\pimport{\Phi}{e}\\
-%%\mvlbl{tycon defs} & 
-%%\Phi & ::= & \keyw{tycon}~\fvar{tc}[\kappa]~ \{\keyw{rep}=\st~\keyw{in}~\omega\} ~|~ \Phi; \Phi\\
-%%\mvlbl{op defs} & \omega & ::= & \keyw{analit}[\kappa]=\st; \keyw{syntarg}[\kappa]=\st; \keyw{anabranch}[\kappa]=\st\\
-%%\end{array}
-%%$
-%%\\
-%%~\\
-%%%\begin{tabular}{ll}
-%%%$\begin{array}[t]{l}
-%%%\mvlbl{static terms}\\
-%%%\st ::= \sx{x}\\
-%%%\slam{\kappa}{x}{\st}\\
-%%%\sap{\st}{\st}\\
-%%%...\\
-%%%\spair{\st}{\st}\\
-%%%\sletpair{\st}{x}{y}{\st}\\
-%%%\snil\\
-%%%\scons{\st}{\st}\\
-%%%\slistrec{\st}{\st}{x}{y}{z}{\st}\\
-%%%\sstr{s}\\
-%%%\sstrrec{\st}{\st}{x}{\st}\\
-%%%\sty{tc}{\st}\\
-%%%\stycase{tc}{\st}{x}{\st}{\st}\\
-%%%\sarg{n}\\
-%%%\sanaarg{\st}{\st}{x}{y}{\st}\\
-%%%\ssynarg{\st}{x}{\st}\\
-%%%\strtm{\qitm}\\
-%%%\strty{\qity}\\
-%%%\styeq{\st}{\st}{\st}{\st}\\
-%%%\sraise\\
-%%%\sval\\
-%%%\end{array}$~$\begin{array}[t]{l}
-%%%\mvlbl{static values}\\
-%%%\sval ::= \\
-%%%\slam{\kappa}{x}{\st}\\
-%%%...\\
-%%%\spair{\sval}{\sval}\\
-%%%\snil\\
-%%%\scons{\sval}{\sval}\\
-%%%\sstr{s}\\
-%%%\sty{tc}{\sval}\\
-%%%\sarg{n}\\
-%%%\saitm{\hitm}\\
-%%%\saity{\hity}\end{array}$
-%%%$\begin{array}[t]{l}
-%%%\\
-%%%\mvlbl{tycon names}\\
-%%%\fvar{tc}\\
-%%%%\text{variables}\\
-%%%%f, x, y, z\\
-%%%%\text{static variables}\\
-%%%%x, y, z\\
-%%%\\
-%%%\\
-%%%\mvlbl{strings}\\
-%%%s\\
-%%%\text{nats}~~
-%%%n
-%%%\end{array}
-%%%$
-%%%$\begin{array}[t]{l}
-%%%\mvlbl{quoted internal types}~~ \qity ::= \\
-%%%%\begin{array}[t]{lcl}
-%%%%\qity & ::= & 
-%%%\qtparr{\qity}{\qity}\pipe
-%%%\qtprod{\qity}{\qity}\pipe
-%%%\qtlist{\qity}\pipe
-%%%\qtstr%\\ & \pipe & 
-%%%\pipe \qtrepof{\st} \pipe
-%%%\qtuq{\st}%\end{array}\\
-%%%\\
-%%%\mvlbl{abstract internal types}~~\hity ::= \\
-%%%%\begin{array}[t]{lcl}
-%%%% & ::= & 
-%%%\htparr{\hity}{\hity}\pipe
-%%%\htprod{\hity}{\hity}\pipe
-%%%1 \pipe + \pipe \mu \pipe 
-%%%%\htlist{\hity}\pipe
-%%%%\htstr\pipe%\\ & \pipe & 
-%%%\htrepof{\sval}
-%%%\\
-%%%\mvlbl{internal types}~~\ity ::= \\
-%%%%\begin{array}[t]{lcl}
-%%%%\ity & ::= & 
-%%%\itparr{\ity}{\ity}\pipe
-%%%\itprod{\ity}{\ity}\pipe
-%%%\itlist{\ity}\pipe
-%%%\itstr%\end{array}\\
-%%%%\end{array}
-%%%%\begin{array}[t]{lcll}
-%%%%\Delta & ::= & \cdot \pipe \Delta, \sx{x}:\kappa & \mvlbl{kinding ctx}\\
-%%%%\Upsilon & ::= & \cdot \pipe \Upsilon, x\Rightarrow\sval & \mvlbl{typing ctx}\\
-%%%%\mathcal{C} & ::= & \cdot \pipe \es;\Upsilon;\Phi & \mvlbl{current op}\\
-%%%%\mvlbl{abstract internal typing context}~\Psi ::= ...\\
-%%%%\Xi & ::= & \cdot \pipe \Xi, \fvar{tc} & \mvlbl{tycon list}\\
-%%%%\es & ::= & \cdot \pipe e; \es & \mvlbl{term lists}\\
-%%%%\qs & ::= & \cdot \pipe \qitm; \qs\\
-%%%%\hs & ::= & \cdot \pipe \hitm; \hs\\
-%%%%\end{array}
-%%%\end{array}$\\
-%%%\end{tabular}~\\
-%%\begin{tabular}{>{$}l<{$}>{$}l<{$}|>{$}l<{$}|>{$}l<{$}>{$}l<{$}|>{$}l<{$}>{$}l<{$}}
-%%\multicolumn{2}{l|}{\mvlbl{kinds \& static terms}} & \mvlbl{external terms} & \multicolumn{2}{l|}{\mvlbl{translational types \& terms}} & \multicolumn{2}{l}{\mvlbl{internal types \& terms}}\\
-%%\kappa ::= & \st ::= & e ::= & \hity ::= & \hitm ::= & \ity ::= & \itm ::=\\
-%%k & \sx{x} & x & t & x & t & x\\
-%%\kty & \sty{tc}{\m}{\st} & \elet{e}{x}{e} & & \hlet{\hitm}{x}{\hitm} & & \ilet{\itm}{x}{\itm}\\
-%%& \m & \\
-%%& \stycase{tc}{\st}{x}{\st}{\st} & \easc{\st}{e} & & \hasc{\hity}{\hitm} & & \iasc{\ity}{\itm}\\
-%%& \styeq{\st}{\st}{\st}{\st} & \elit{\st}{\es} & & \hlit{\st}{\hs}\\
-%%\karrow{\kappa}{\kappa} & \slam{\kappa}{x}{\st} & \etarg{\st}{e}{\es} & & \htarg{\st}{h}{\hs} & \\
-%% & \sap{\st}{\st} & \ebranch{\st}{e}{\es} & & \hbranch{\st}{h}{\hs}\\
-%%\kunit & \striv & \elamrec{f}{x}{e} & \htparr{\hity}{\hity} & \hlamrec{f}{x}{\hitm} & \itparr{\ity}{\ity} & \ilamrec{f}{x}{\itm}\\
-%%\kprod{\kappa}{\kappa} & \spair{\kappa}{\kappa} & & & \hap{\hitm}{\hitm} & & \iap{\itm}{\itm}\\
-%%& \sletpair{\st}{x}{y}{\st} & & \htunit & \htriv & \itunit & \itriv\\
-%%\ksum{\kappa}{\kappa} & \sinl{\kappa}{\st} & & \htprod{\hity}{\hity} & \hpair{\hitm}{\hitm} & \itprod{\ity}{\ity} & \ipair{\itm}{\itm}\\
-%%& \sinr{\kappa}{\st} & & & \hletpair{\hitm}{x}{y}{\hitm} & & \iletpair{\itm}{x}{y}{\itm}\\
-%%& \scase{\st}{x}{\st}{x}{\st} & & \htsum{\hity}{\hity} & \hinl{\hity}{\hitm} & \itsum{\ity}{\ity} & \iinl{\ity}{\itm}\\
-%%\kind{k}{\kappa} & \sfold{k}{\kappa}{\st} & & & \hinr{\hity}{\hitm} & & \iinr{\ity}{\itm}\\
-%%& \srec{\kappa}{\st}{x}{\st} & & & \hcase{\hitm}{x}{\hitm}{x}{\hitm} & & \icase{\itm}{x}{\itm}{x}{\itm}\\
-%%\ktrty & \strty{\hity} & & \htrec{t}{\hity}\\
-%%\ktrtm & \strtm{\hitm}\\
-%%& \sana{n}{\st}{\st}\\
-%%& \ssyn{n}{\st}\\
-%%& \sraise & ~& & &2
-%%\end{tabular}\\
-%%$\begin{array}[t]{l}
-%%\mvlbl{external terms}\\
-%%e ::= x\\
-%%\elet{e}{x}{e}\\
-%%\easc{\st}{e}\\
-%%\elit{\st}{\es}\\
-%%\etarg{\st}{e}{\es}\\
-%%\ebranch{\st}{e}{\es}\\
-%%\elamrec{f}{x}{e}\\
-%%%\qderivate{n}{\sval}{\hitm}\\
-%%\end{array}$~
-%%$\begin{array}[t]{l}
-%%%~\\
-%%\mvlbl{static terms}\\
-%%\st ::= \sx{x} \\
-%%\sty{tc}{t}{\st} \\
-%% \\
-%%\slam{\kappa}{x}{\st} \\ 
-%%\sap{\st}{\st} \\
-%%\striv\\
-%%\spair{\st}{\st}\\
-%%\sletpair{\st}{x}{y}{\st}\\
-%%\sinl{\kappa}{\st}\\
-%%\sinr{\kappa}{\st}\\
-%%\scase{\st}{x}{\st}{x}{\st}\\
-%%~\\
-%%~\\
-%%  \strtm{\hitm} \\
-%%  \strty{\hity}\\
-%%\sana{n}{\st}{\st} \\
-%% \ssyn{n}{\st}  \\
-%%% & & \pipe & \styeq{\st}{\st}{\st}{\st}\pipe\sraise\\
-%%%\mvlbl{kinds} & \kappa & ::= & k \pipe \karrow{\kappa}{\kappa} \pipe \kunit \pipe \kprod{\kappa}{\kappa} \pipe \ksum{\kappa}{\kappa} \pipe \kind{k}{\kappa} \\
-%%% & & \pipe & \kty \pipe \ktrtm \pipe \ktrty\\
-%%%~\\
-%%\end{array}$~
-%%$\begin{array}[t]{l}
-%%\mvlbl{kinds}\\
-%%\kappa ::= k\\
-%%\kty\\
-%%~\\
-%%\karrow{\kappa}{\kappa}\\
-%%~\\
-%%\kunit\\
-%%\kprod{\kappa}{\kappa}\\
-%%~\\
-%%\ksum{\kappa}{\kappa}\\
-%%~\\
-%%~\\
-%%\kind{k}{\kappa}\\
-%%~\\
-%%\ktrtm\\
-%%\ktrty
-%%\end{array}$~
-%%$\begin{array}[t]{l}
-%%\mvlbl{translational terms}\\
-%%\hitm ::= x\\
-%%\hlet{\hitm}{x}{\hitm}\\
-%%\hasc{\hity}{\hitm}\\
-%%\hlit{\sval}{\hs}\\
-%%\htarg{\sval}{\hitm}{\hs}\\
-%%\hbranch{\st}{\hitm}{\hs}\\
-%%\hlamrec{f}{x}{\hitm}\\
-%%\hap{\hitm}{\hitm}\\
-%%\hpair{\hitm}{\hitm}\\
-%%\hletpair{\hitm}{x}{y}{\hitm}\\
-%%\hnil\\
-%%\hcons{\hitm}{\hitm}\\
-%%%\hlistrec{\hitm}{\hitm}{x}{y}{z}{\hitm}\\
-%%\hstr{s}\\
-%%\hstrrec{\hitm}{\hitm}{x}{\hitm}\\
-%%\quq{\st}\\
-%%\hderivate{n}{\sval}{\hitm}\\
-%%\end{array}$~
-%%$\begin{array}[t]{l}
-%%\mvlbl{translational types}\\
-%%%\mvlbl{types}\\
-%%\hity ::= t\\
-%%~\\
-%%~\\
-%%~\\
-%%~\\
-%%~\\
-%%arrow\\
-%%~\\
-%%1\\
-%%prod\\
-%%~\\
-%%sum\\
-%%~\\
-%%~\\
-%%rec\\
-%%~\\
-%%uq\\
-%%repoft
-%%\end{array}$~
-%%$\begin{array}[t]{l}
-%%\mvlbl{internal terms}\\
-%%\dot{\iota} ::= x\\
-%%  \ilet{\itm}{x}{\itm}\\
-%%\iasc{\ity}{\itm}\\
-%%~\\
-%%~\\
-%%~\vspace{3px}\\
-%%\ilamrec{f}{x}{\itm}\\
-%%\iap{\itm}{\itm}\\
-%%\ipair{\itm}{\itm}\\
-%%\iletpair{\itm}{x}{y}{\itm}\\
-%%\inil\\
-%%\icons{\itm}{\itm}\\
-%%\ilistrec{\itm}{\itm}{x}{y}{z}{\itm}\\
-%%\istr{s}\\
-%%\istrrec{\itm}{\itm}{x}{\itm}
-%%\end{array}$~
-%%$\begin{array}[t]{l}
-%%\mvlbl{internal types}\\
-%%\hity ::= t\\
-%%~\\
-%%~\\
-%%~\\
-%%~\\
-%%~\\
-%%arrow\\
-%%~\\
-%%1\\
-%%prod\\
-%%~\\
-%%sum\\
-%%~\\
-%%~\\
-%%rec\\
-%%\end{array}$
-%%\caption{Test}
-%%\end{figure*}
-%%\begin{figure*}
-%%\small
-%%\fbox{$\esynX{e}{\sval}{\hitm}$}~
-%%\fbox{$\eanaX{e}{\sval}{\hitm}$}~
-%%%\fbox{$\etyerrX{e}$}~
-%%%\fbox{$\efragerrX{e}$}
-%%\begin{mathpar}
-%%\inferrule[syn-var]{
-%%	\tctxbind{x}{\sval} \in \Upsilon
-%%}{	
-%%	\esynX{x}{\sval}{x}
-%%}
-%%
-%%%\inferrule[tyerr-var]{
-%%%	\tctxbind{x}{\sval} \notin \Upsilon
-%%%}{
-%%%	\etyerrX{x}
-%%%}
-%%%
-%%\inferrule[syn-let]{
-%%   \esynX{e_1}{\sval_1}{\hitm_1}\\
-%%   \esyn{\Upsilon, \tctxbind{x}{\sval_1}}{\Phi}{e_2}{\sval_2}{\hitm_2}
-%%}{
-%%	\esynX{\elet{e_1}{x}{e_2}}{\sval_2}{\hlet{\hitm_1}{x}{\hitm_2}}
-%%}
-%%
-%%\inferrule[ana-let]{
-%%   \esynX{e_1}{\sval_1}{\hitm_1}\\
-%%   \eana{\Upsilon, \tctxbind{x}{\sval_1}}{\Phi}{e_2}{\sval_2}{\hitm_2}
-%%}{
-%%	\eanaX{\elet{e_1}{x}{e_2}}{\sval_2}{\hlet{\hitm_1}{x}{\hitm_2}}
-%%}
-%%
-%%\inferrule[syn-slet]{
-%%	\skind{\cdot}{\Phi}{\st}{\kappa}\\
-%%	\esynX{[\st/x]e}{\sval}{\hitm}
-%%}{
-%%	\esynX{\eslet{\st}{x}{e}}{\sval}{\hitm}
-%%}
-%%
-%%\inferrule[ana-slet]{
-%%	\skind{\cdot}{\Phi}{\st}{\kappa}\\
-%%	\eanaX{[\st/x]e}{\sval}{\hitm}
-%%}{
-%%	\eanaX{\eslet{\st}{x}{e}}{\sval}{\hitm}
-%%}
-%%
-%%\inferrule[syn-asc]{
-%%	\skind{\cdot}{\Phi}{\st}{\kty}\\
-%%	\sevalN{\st}{\sval}\\
-%%	\eanaX{e}{\sval}{\hitm}
-%%}{
-%%	\esynX{\easc{\st}{e}}{\sval}{\hitm}
-%%}
-%%
-%%\inferrule[ana-lamrec]{
-%%	\Phi_0 \in \Phi\\
-%%	\eana{\Upsilon, \tctxbind{f}{\sty{parrow}{t}{\spair{\sval_1}{\sval_2}}}, \tctxbind{x}{\sval_1}}{\Phi}{e}{\sval_2}{\hitm}
-%%}{
-%%	\eanaX{\elamrec{f}{x}{e}}{\sty{parrow}{t}{\spair{\sval_1}{\sval_2}}}{\hlamrec{f}{x}{\hitm}}
-%%}
-%%
-%%\inferrule[ana-lit]{
-%%  \fvar{tc}[\kx{tyidx}] \{\keyw{analit}[\kx{tmidx}]=\stx{analit}; \cdots\} \in \Phi\\\\
-%%  \skind{\cdot}{\Phi}{\svalx{tmidx}}{\kx{tmidx}}\\
-%%  \args\\\\
-%%  \seval{\stx{analit}~\sty{tc}{t}{\svalx{tyidx}}~\svalx{tmidx}~\svalx{args}}{\Upsilon;\Phi;\Delta}{\saitm{\hitm}}\\\\
-%%  check
-%%}{
-%%  \eanaX{\elit{\svalx{tmidx}}{\es}}{\sty{tc}{t}{\svalx{tyidx}}}{\hitm}
-%%}
-%%\end{mathpar}
-%%\caption{Abstract Active Typechecking and Translation}
-%%\end{figure*}
-%%\inferrule[import]{
-%%	\vdash \Phi\\
-%%	\Upsilon \vdash_\Phi e \Rightarrow \sty{tc}{\sval} \leadsto \iota
-%%}{
-%%	\keyw{import}~\Phi~\keyw{in}~e : \sty{tc}{\sval} \leadsto \iota
-%%}
-%
-%%
-%%\begin{figure}[t]
-%%\begin{lstlisting}
-%%tycon Nat of 1 with  
-%%    schema $\lambda$idx:1.$\blacktriangledown$($\mathbb{Z}$) 
-%%    opcon Z of 1 ($\lambda$idx:1.$\lambda$args:list[Elab].is_empty args $\llbracket$$\triangledown$(0) as Nat[()]$\rrbracket$)
-%%    opcon S of 1 ($\lambda$idx:1.$\lambda$args:list[Elab].pop_final args $\lambda$x:ITm.$\lambda$ty:$\star$. 
-%%      checktype ty Nat[()] $\llbracket$$\triangledown$($\vartriangle$(x)+1) as Nat[()]$\rrbracket$)
-%%    opcon Rec of 1 ($\lambda$idx:1.$\lambda$args:list[Elab].
-%%      pop args $\lambda$x1:ITm.$\lambda$ty1:$\star$.$\lambda$args':list[Elab].
-%%      pop args' $\lambda$x2:ITm.$\lambda$ty2:$\star$.$\lambda$args'':list[Elab].
-%%      pop_final args'' $\lambda$x3:ITm.$\lambda$ty3:$\star$.
-%%      check_type t1 Nat[()] (
-%%      check_type t3 Arrow[(Nat[()], Arrow[(t2, t2)])] 
-%%      	$\llbracket$$\triangledown$((fix f:$\mathbb{Z} \rightarrow$ rep(t2) is $\lambda$x:$\mathbb{Z}$.
-%%	        if x = 0 then $\vartriangle$(x2) else $\vartriangle$(x3) (x - 1) (f (x - 1))) $\vartriangle$(x1)) as t2$\rrbracket$))
-%%end
-%%\end{lstlisting}
-%%\caption{An implementation of primitive natural numbers as internal integers in @$\lambda$.}
-%%\label{nat-atlam}
-%%\end{figure}
-%
-%The syntax of Core @$\lambda$ is given in Fig.
-\ref{grammar}.
-An example of a program defining type and operator constructors that admit an active embedding of G\"odel's \textbf{T} into @$\lambda$ is given in Fig.
-\ref{nat}.
-We will discuss its semantics and how precisely that embedding, seen being used starting on line 16, works as we go on.
-Natural numbers can, of course, be strongly embedded in existing languages, with a similar usage and asymptotic performance profile (up to function call overhead as an abstract type, for example).
-We will provide more sophisticated examples where this is less feasible later on (and note that type abstraction itself .....
-in Sec.
-\ref{limitations}).\todo{if i don't have time/space, take this out} %It consists of:
-%%a type constructor declaration, $\fvar{Nat}$, indexed trivially, together with three operator constructors, also all indexed trivially, that implement the standard introductory forms for natural numbers as well as the recursor operator (as in G\"odel's T \cite{pfpl}).
-Following the type constructor declaration, we apply $\fvar{Nat}$ with the trivial index, $\tunit$, to form the type $\tvar{nat}$.
-Finally, we write an external term that uses the operators associated with $\fvar{nat}$ and the built-in constructor $\fvar{Parr}$, governing partial functions, to define an addition function and compute the addition of the natural numbers  two and two.
-We will introduce a more convenient concrete syntax in later portions of this thesis; for now we will restrict ourselves to the abstract syntax so that this example can directly aid in understanding the semantics.
-%
-%\subsection{Overview}\label{programs}
-%%\begin{figure}[t]
-%%\small
-%%$$\begin{array}{rccl}	
-%%\textbf{programs} & \rho & ::= & \pfam{\familyDf}{\progsort}
-%%\\& &  \pipe & 
-%%%\pipe \pdef{t}{\kappa}{\tau}{\progsort} 
-%% e\\
-%%		&	\theta	&	::= &	\tops{op}{\kappaidx}{i}{a}{\taut{def}} \pipe 
-%%												\topp{\theta}{\theta}\\
-%%\\
-%%\textbf{external terms} 				&	e	&	::=	&	\evar{x} \pipe 
-%%%														\efix{x}{\tau}{e} \pipe 
-%%														%\elam{\evar{x}}{\tau}{e} \pipe 
-%%														\eop{Tycon}{op}{
-%%															\tauidx
-%%														}{
-%%  												    		\splat{e}{1}{n}
-%%														} \\
-%%									& 		&		& 	\\
-%%
-%%
-%%\textbf{internal terms} 				& 	\iota	&	::=	&	\evar{x} \pipe 
-%%												\ifix{\evar{x}}{\sigma}{\iota} \pipe
-%%												\ilam{\evar{x}}{\sigma}{\iota} \pipe 
-%%												\iapp{\iota_{1}}{\iota_{2}} 
-%%\\\text{integers}&&\pipe&												
-%%												\iintlit \pipe \iop{\iota_{1}}{\iota_{2}} \pipe \iIfEq{\iota_{1}}{\iota_{2}}{\dint}{\iota_{3}}{\iota_{4}} \\
-%%\text{products}										& & \pipe & 
-%%												\iunit \pipe
-%%												\ipair{\iota_{1}}{\iota_{2}} \pipe 
-%%												\ifst{\iota} \pipe
-%%												\isnd{\iota} 
-%%\\\text{sums}&&\pipe&												
-%%												 \iinl{\sigma_2}{\iota_1} \pipe \iinr{\sigma_1}{\iota_2} \\&&\pipe& \icase{e}{x}{e_1}{x}{e_2} \\
-%%												
-%%%\text{deabstracted}& \iota & ::= & \mathcal{G}[\iota, \sigma]\\
-%%\textbf{internal types}			&	\sigma	&	::=	&    \darrow{\sigma_1}{\sigma_2} \pipe \dint \pipe \dunit \pipe \dpair{\sigma_1}{\sigma_2} \pipe \dsum{\sigma_1}{\sigma_2}  
-%%												
-%%\\
-%%\\
-%%							
-%%\hspace{-5pt}\textbf{type-level terms} 	& \tau 	& ::= 	& 	\tvar{t} \pipe 
-%%														\tlam{t}{\kappa}{\tau} \pipe 
-%%														\tapp{\tau_1}{\tau_2} \pipe \iintlit \pipe \iop{\tau_1}{\tau_2} \pipe \tlabel{label}
-%%\\\text{lists}&&\pipe&											
-%%														\tnil{\kappa} \pipe \tcons{\tau_1}{\tau_2} \pipe 
-%%									                     \tfold{\tau_1}{\tau_2}{h}{t}{r}{\tau_3}
-%%														\\
-%%												
-%%\text{products}	 			& 		& \pipe	& 	  \tunit \pipe 
-%%														\tpair{\tau_{1}}{\tau_{2}} \pipe 
-%%														\tfst{\tau} \pipe 
-%%														\tsnd{\tau} 
-%%														\\	
-%%\text{sums}									&       & \pipe & \tinl{\kappa_2}{\tau_1} \pipe \tinr{\kappa_1}{\tau_2} \\&&\pipe& \tsumcase{\tau}{t}{\tau_1}{t}{\tau_2} \\
-%%						\text{types} 						& 		& \pipe	& 	\ttypestd \\&&\pipe& \tfamcase{\tau}{Tycon}{x}{\tau_1}{\tau_2}\\
-%%\text{equality}  & & \pipe & 					\tifeq{\tau_{1}}{\tau_{2}}{\kappa}{\tau_{3}}{\tau_{4}} 
-%%														\\													%				& & \pipe & \tfamcase{\tau}{Fam}{x}{\tau_1}{\tau_2}\\
-%%																								
-%%\text{derivates} 				& 		 & 	\pipe	&	\tden{{\bar\iota}}{\tau} \pipe \tden{\ibar}{\tau}^{\checkmark} \pipe \ttypeof{\tau} \\% \tdencase{\tau}{x}{t}{\tau_1}{\tau_2}\\
-%% %& & \pipe & 
-%%%														\tdencase{\tau}{y}{x}{\tau_1}{\tau_2}
-%%%														 \\
-%%
-%%\text{rep types}		&		&	\pipe	&	\titype{\bar \sigma} \\
-%%% &  & \pipe & \tvalof{\tau_1}{\tau_2} \pipe \iup{\tau} \\
-%%% 												\trepof{\tau} \pipe \dup{\tau}\\
-%%\text{translational IL}	& \bar{\iota} & ::= & x \pipe \ifix{x}{\bar \sigma}{\bar \iota} 
-%%	%\pipe \ilam{x}{\bar \sigma}{\bar \iota} \pipe \iapp{\bar \iota_1}{\bar \iota_2} 
-%%	\pipe \cdots \pipe \itransof{\tau} \\						
-%% & \bar{\sigma} & ::= & \darrow{\bar \sigma_1}{\bar \sigma_2} \pipe \cdots \pipe \dup{\tau} \pipe \trepof{\tau} \\
-%%%\text{abstracted} & \sabs & ::= & \darrow{\sabs_1}{\sabs_2} \pipe \cdots \pipe \sabsrep{\tau}
-%%											\\
-%%\textbf{kinds} 					& \kappa	&	::=	&	\karrow{\kappa_1}{\kappa_2} \pipe \kint \pipe
-%%											    \klabel \pipe
-%%											    \klist{\kappa} \pipe
-%%												\kunit \pipe 
-%%												\kpair{\kappa_{1}}{\kappa_{2}} \\
-%%												&&\pipe&
-%%												\ksum{\kappa_1}{\kappa_2} \pipe
-%%												\kTypeBlur \pipe \kDen \pipe 
-%%												\kIType								
-%%%\textbf{ops signature}			& \Theta	&	::=	&	\kOpEmpty \pipe \kOp{\Theta}{op}{\kappai}\\
-%%%											 							&		&		&	\\
-%%\end{array}$$
-%%%\vspace{-10pt}
-%%\caption{\small Syntax of Core \atlam.
-Here, $x$ ranges over external and internal language variables, $\tvar{t}$ ranges over type-level variables, $\fvar{Tycon}$ ranges over type constructor names, $\opvar{op}$ ranges over operator constructor names, $\iintlit$ ranges over integer literals, $\tlabel{label}$ ranges over label literals (see text) and $\oplus$ ranges over standard total binary operations (e.g.
-addition, comparison).
-%%\label{grammar}}
-%%\end{figure}
-%%\begin{figure}[t]
-%%\small
-%%\begin{flalign}
-%%\label{natfam}&\family{Nat}{\kunit}{\\
-%%\label{z}&\quad\tops{z}{\kunit}{i}{a}{\tlam{i}{\kunit}{\tlam{a}{\klist{\kDen}}{
-%%	\tapp{\tapp{\tvar{arity0}}{\tvar{a}}}{\\&\quad\quad
-%%		\tden{0}{\ttype{Nat}{\tunit}}
-%%	}}}};\\
-%%\label{s}&\quad\tops{s}{\kunit}{i}{a}{\tlam{i}{\kunit}{\tlam{a}{\klist{\kDen}}{\tvar{arity1}~\tvar{a}~\tlam{d}{\kDen}{\\
-%%& \quad\quad \tvar{ifeq}~\ttypeof{\tvar{d}}~\ttype{Nat}{\tunit}\\
-%%& \quad\quad\quad \tden{\itransof{\tvar{d}} + 1}{\ttype{Nat}{\tunit}}
-%%	}}}};\\
-%%\label{rec}&\quad\tops{rec}{\kunit}{i}{args}{\tlam{i}{\kunit}{\tlam{a}{\klist{\kDen}}{\tvar{arity3}~\tvar{a}~\tlam{d1}{\kDen}{\tlam{d2}{\kDen}{\tlam{d3}{\kDen}{\\
-%%&\quad\quad \tvar{ifeq}~\ttypeof{\tvar{d1}}~\ttype{Nat}{\tunit}\\
-%%&\quad\quad {\sf let}~\tvar{t2} = \ttypeof{\tvar{d2}}~{\sf in}\\
-%%&\quad\quad \tvar{ifeq}~\ttypeof{\tvar{d3}}~\ttype{Arrow}{(\ttype{Nat}{\tunit}, \ttype{Arrow}{(\tvar{t2}, \tvar{t2})})}\\
-%%&\quad\quad\quad \tden{\iapp{(\ifix{f}{\darrow{\dint}{\trepof{\tvar{t2}}}}{\ilam{x}{\dint}{\\
-%%		\label{lastop}&\quad\quad\quad\quad\quad \iIfEq{x}{0}{\dint}{\itransof{\tvar{d2}}}{\\
-%%		&\quad\quad\quad\quad\quad\quad\iapp{\iapp{\itransof{\tvar{d3}}}{(x-1)}}{(\iapp{f}{(x-1)})}}}}\\
-%%		&\quad\quad\quad\quad)}{\itransof{\tvar{d1}}}}{\tvar{t2}}
-%%}}}
-%%}}}
-%%\\
-%%&}{XXX}{i}{\tlam{i}{\kunit}{\titype{\dint}}};\\
-%%\label{nattype}&{\sf let}~\tvar{nat} = \ttype{Nat}{\tunit}~{\sf in}\\
-%%& \elet{two}{\eop{Nat}{s}{\tunit}{\eop{Nat}{s}{\tunit}{\eop{Nat}{z}{\tunit}{}}}}{\\
-%%& \elet{plus}{
-%%	\elam{x}{\tvar{nat}}{
-%%		\elam{y}{\tvar{nat}}{
-%%			\\&\quad \eop{Nat}{rec}{\tunit}{x; y; \elam{p}{\tvar{nat}}{\elam{r}{\tvar{nat}}{
-%%				\eop{Nat}{s}{\tunit}{r}
-%%			}}}
-%%		}
-%%	}
-%%}{
-%%& \\
-%%& \eop{Arrow}{ap}{\tunit}{plus; \eop{Arrow}{ap}{\tunit}{two; two}}
-%%}}
-%%%{\\&\elet{two}{\tvar{nat}}{\eop{Nat}{s}{\tunit}{\eop{Nat}{s}{\tunit}{\eop{Nat}{z}{\tunit}{ }}}}{\eapp{\eapp{plus}{two}}{two}}}
-%%%}
-%%%\elam{x}{\tvar{nat}}{\elam{y}{\tvar{nat}}{\\
-%%	%&\quad \eop{Nat}{rec}{\tunit}{x; y; \elam{p}{\tvar{nat}}{\elam{r}{\tvar{nat}}{
-%%	%\eop{Nat}{s}{\tunit}{r}
-%%	%}}}}
-%%\end{flalign}
-%%%\vspace{-15px}
-%%\caption{\small An embedding of G\"odel's $\mathbf{T}$ in \atlam, used to define $plus$ and $two$ and calculate $plus~two~two$.
-Type constructors (line 1) and types (line 16) are discussed in Sec.
-\ref{types}.
-Schemas (line 1) and operator constructors (lines 2-14, used on lines 17-20) are discussed starting in Sec.
-\ref{opcons}.
-We use \textsf{let} to bind both external and type-level variables as needed for clarity of presentation.
-We will discuss further syntactic issues in Sec.
-\ref{expanded-syntax}.}
-%%\label{nat}
-%%\vspace{-10pt}
-%%\end{figure}
-%A \emph{program}, $\rho$, consists of a series of static declarations (in the core, only constructor declarations) followed by an external term, $e$.
-The syntax for external terms contains three forms: variables, $\lambda$ terms, and a form for invoking user-defined operators, which we will discuss below.
-Compiling a program consists of first \emph{kind checking} it then typechecking the external term and {translating} it to a term, $\iota$, in the {typed internal language}.
-These correspond to the premises of the \emph{central compilation judgement} $\pkcompiles{\rho}{\iota}$:
-%\[
-%\inferrule[p-compiles]{
-%	\emptyset \vdash_{\fvalCtx_0} \rho\\
-%%	\progOK{\emptyset}{\fvalCtx_0}{\rho}\\
-%	\pcompiles{\fvalCtx_0}{\rho}{\iota}
-%}{\pkcompiles{\rho}{\iota}}
-%\]
-%This anchors our exposition; we will describe how it is derived (i.e. how to write a compiler for @$\lambda$) in the following sections. 
-%
-%The key judgement in the calculus is the \emph{active typing judgement}  (Fig.
-\ref{att}, which we describe starting in Sec.
-\ref{opcons}).
-It relates an external term, $e$, to a {type}, $\tau$, called its \emph{type assignment}, and an internal term, $\iota$, called its \emph{translation}, under \emph{typing context} $\Upsilon$ and \emph{constructor context} $\Phi$:
-%\[\ecompilesX{e}{\tau}{\iota}\]
-%
-%The typing context $\Upsilon$ maps variables to types in essentially the conventional way (\cite{pfpl} contains the necessary background for this paper).
-The constructor context tracks user-defined type and operator constructors.
-% Weakening and exchange of the constructor context closely related to the issue of conservativity that we will return to.
-Each constructor is identified by name, so there is no analog to contraction.
-%There is no separate operational semantics for the external language.
-Instead, the dynamic behavior of an external term is determined by its translation to the internal language, which has a more conventional operational semantics.
-This form of semantics can be seen as lifting into the language specification the first stage of a type-directed compiler like the TIL compiler for Standard ML \cite{tarditi+:til-OLD} and has some parallels to the Harper-Stone semantics for Standard ML, where external terms were also given meaning by elaboration from the EL to an IL \cite{Harper00atype-theoretic}.
-%
-%
-%In @$\lambda$, the internal language (IL) provides partial functions (via the generic fixpoint operator of Plotkin's PCF), simple product and sum types and a base type of integers (to make our example interesting and as a nod toward speed on contemporary machines).
-In practice, the internal language could be any typed  language with a specification for which type safety and decidability of typechecking have been satisfyingly determined.
-The internal type system serves as a ``floor'': guarantees that must hold for \emph{all} well-typed terms, independent of the constructor context (e.g.
-that out-of-bounds access to memory never occurs), must be maintained by the internal type system.
-User-defined constructors can (only) enforce invariants stronger than those the internal type system maintains.
-Performance is also ultimately limited by the internal language and downstream compilation stages that we do not here consider (safe compiler extension has been discussed in previous work, e.g.
-\cite{conf/pldi/TatlockL10}).
-%
-%\subsection{Type Constructors and Types}\label{types}
-%\begin{figure}[t]
-%\small
-%$\fbox{\inferrule{}{\progOKX{\progsort}}}$
-%~~~$\tvarCtx ::= \emptyctx \pipe \tvarCtxX{t}{\kappa}$
-%~~~$\fvalCtx ::= \emptyctx \pipe \fvalCtxX{\fvalDf}$
-%%$\fCtx ::= \Sigma_0	 \pipe \fvalCtxX$
-%\begin{mathpar}
-%\inferrule[k-tycon]{
-%	\fvar{Tycon} \notin \text{dom}(\fvalCtx)\\
-%	\kEq{\kappaidx}\\
-%	\tKind{\tvarCtx}{\fvalCtx}{\taurep}{\karrow{\kappaidx}{\kIType}}\\
-%	\opType{\tvarCtx}{\fvalCtxX{\fvalDf}}{\theta}{\Theta}\\
-%	\progOK{\tvarCtx}{\fvalCtxX{\fvalDf}}{\rho}
-%}{
-%	\progOKX{\pfam{\familyDf}{\rho}}
-%}
-%
-%%\inferrule[def-kinding]{
-%%	\tKindX{\tau}{\kappa}\\
-%%	\progOK{\tvarCtxX{t}{\kappa}}{\fvalCtx}{\rho}
-%%}{
-%%	\progOKX{\pdef{t}{\kappa}{\tau}{\rho}}
-%%}
-%%
-%\inferrule[k-e-prog]{
-%	\exprOK{\tvarCtx}{\fvalCtx}{e}
-%}{
-%	\progOKX{e}
-%}
-%\end{mathpar}
-%$\fbox{$\opType{\tvarCtx}{\fvalCtx}{\theta}{\Theta}$}$
-%\begin{mathpar}
-%\inferrule[k-opcon]{
-%	\tKind{\tvarCtx}{\fvalCtx}{\taudef}{\karrow{\kappaidx}{\karrow{\klist{\kDen}}{(\ksum{\kDen}{\kunit})}}}
-%}{
-%	\opType{\tvarCtx}{\fvalCtx}{{\tops{op}{\kappaidx}{i}{a}{\taudef}}}{
-%	\kOpS{op}{\kappaidx}}
-%}
-%
-%\inferrule[k-opcons]{
-%	\opType{\tvarCtx}{\fvalCtx}{\theta_1}{\Theta_1}\\
-%	\opType{\tvarCtx}{\fvalCtx}{\theta_2}{\Theta_2}\\\\
-%	\text{dom}(\theta_1) \cap \text{dom}(\theta_2) = \emptyset
-%}{
-%	\opType{\tvarCtx}{\fvalCtx}{\theta_1; \theta_2}{\Theta_1, \Theta_2}
-%}
-%\end{mathpar}
-%$\fbox{$\exprOKX{e}$}$
-%\begin{mathpar}
-%\inferrule[k-e-var]{ }{
-%	\exprOK{\tvarCtx}{\fvalCtx}{\evar{x}}
-%}
-%
-%%\inferrule[k-e-lam]{
-%%	\tKindX{\tau}{\kTypeBlur}\\
-%%	\exprOK{\tvarCtx}{\fvalCtx}{e}
-%%}{
-%%	\exprOKX{\elam{\evar{x}}{\tau}{e}}
-%%}
-%
-%\inferrule[k-e-op]{
-%	\fval{Tycon}{-}{-}{\theta} \in \fvalCtx\\
-%	\tops{op}{\kappaidx}{i}{a}{-} \in \theta\\
-%	\tKindX{\tauidx}{\kappaidx}\\
-%	\exprOKX{e_1}\\
-%	\cdots\\
-%	\exprOKX{e_n}
-%}{
-%	\exprOKX{\eop{Tycon}{op}{\tauidx}{\splat{e}{1}{n}}}
-%}
-%\end{mathpar}
-%$\fbox{$\kEq{\kappa}$}$
-%\begin{mathpar}
-%\inferrule[t-eq]{ }{
-%	\kEq{\kTypeBlur}
-%}
-%
-%\inferrule[i-eq]{ }{
-%	\kEq{\kint}
-%}
-%
-%\inferrule[l-eq]{ }{
-%	\kEq{\klabel}
-%}
-%
-%\inferrule[list-eq]{
-%	\kEq{\kappa}
-%}{
-%	\kEq{\klist{\kappa}}
-%}
-%\\
-%\inferrule[u-eq]{ }{
-%	\kEq{\kunit}
-%}
-%
-%\inferrule[p-eq]{
-%	\kEq{\kappa_1}\\
-%	\kEq{\kappa_2}
-%}{
-%	\kEq{\kpair{\kappa_1}{\kappa_2}}
-%}
-%
-%\inferrule[s-eq]{
-%	\kEq{\kappa_1}\\
-%	\kEq{\kappa_2}
-%}{
-%	\kEq{\ksum{\kappa_1}{\kappa_2}}
-%}
-%\end{mathpar}
-%\caption{\small Kinding for type and operator constructors and external terms, and equality kinds (see text).
-Type-level terms stored in the constructor context are not needed during kinding, indicated using a dash.}
-%\label{kindprog}
-%\vspace{-10pt}
-%\end{figure}
-%\begin{figure}[t]
-%\small
-%%\vspace{-15pt}
-%$\fbox{\inferrule{}{\tKindX{\tau}{\kappa}}}$
-%%~~~$\chi ::= \cdot \pipe \checkmark$
-%\begin{mathpar}
-%\small\inferrule[k-var]{
-%}{
-%  \tKind{\tvarCtxX{t}{\kappa}}{\fvalCtx}{\tvar{t}}{\kappa}
-%}
-%~~~~
-%\inferrule[k-arrow-i]{
-%  \tKind{\tvarCtxX{t}{\kappa_1}}{\fvalCtx}{\tau}{\kappa_2}
-%}{
-%  \tKindX{\tlam{t}{\kappa_1}{\tau}}{\karrow{\kappa_1}{\kappa_2}}
-%}
-%~~~~
-%\inferrule[k-arrow-e]{
-%  \tKindX{\tau_1}{\karrow{\kappa_1}{\kappa_2}}\\\\
-%  \tKindX{\tau_2}{\kappa_1}
-%}{
-%  \tKindX{\tapp{\tau_1}{\tau_2}}{\kappa_2}
-%}
-%
-%\text{\color{gray} (kinding for integers, labels, lists, products and sums also standard)}
-%
-%%%
-%%%\inferrule{ }{
-%%%	\tKindX{\tstr{str}}{\kstr}
-%%%}(\text{str}^I_\tau)
-%%%
-%%%\inferrule{ }{
-%%%	\tKindX{\tunit}{\kunit}
-%%%}(\text{1}^I_\tau)
-%%%
-%%%\inferrule{
-%%%	\tKindX{\tau_1}{\kappa_1}\\
-%%%	\tKindX{\tau_2}{\kappa_2}
-%%%}{
-%%%	\tKindX{\tpair{\tau_1}{\tau_2}}{\kpair{\kappa_1}{\kappa_2}}
-%%%}({\times}^I_\tau)
-%%%
-%%%\inferrule{
-%%%	\tKindX{\tau}{\kpair}{\kappa_1}{\kappa_2}
-%%%}{
-%%%	\tKindX{\tfst{\tau}}{\kappa_1}
-%%%}({\times}^{E1}_\tau)
-%%%
-%%%\inferrule{
-%%%	\tKindX{\tau}{\kpair}{\kappa_1}{\kappa_2}
-%%%}{
-%%%	\tKindX{\tsnd{\tau}}{\kappa_2}
-%%%}({\times}^{E2}_\tau)
-%%%\inferrule{ }{
-%%%\inferrule{
-%%%	\tKindX{\tau_{1}}{\kappa_{1}}\\
-%%%	\tKindX{\tau_{2}}{\kappa_{2}}
-%%%}{
-%%%	\tKindX{\tpair{\tau_{1}}{\tau_{2}}}{\kpair{\kappa_{1}}{\kappa_{2}}}
-%%%}~(\times_\tau)
-%%%
-%%%\inferrule{
-%%%	\tKindX{\tau}{\kpair{\kappa_{1}}{\kappa_{2}}}
-%%%}{
-%%%	\tKindX{\tfst{\tau}}{\kappa_{1}}
-%%%}~(\text{fst}_\tau)
-%%%
-%%%\inferrule{
-%%%	\tKindX{\tau}{\kpair{\kappa_{1}}{\kappa_{2}}}
-%%%}{
-%%%	\tKindX{\tsnd{\tau}}{\kappa_{2}}
-%%%}~(\text{snd}_\tau)
-%%%
-%%% TODO: list
-%\inferrule[k-eq]{
-%	\kEq{\kappa}\\
-%	\tKindX{\tau_1}{\kappa}\\
-%	\tKindX{\tau_2}{\kappa}\\\\
-%	\tKindX{\tau_3}{\kappa'}\\
-%	\tKindX{\tau_4}{\kappa'}
-%}{
-%	\tKindX{\tifeq{\tau_1}{\tau_2}{\kappa}{\tau_3}{\tau_4}}{\kappa'}
-%}
-%
-%\inferrule[k-ty-i]{
-%	\fval{Tycon}{\kappaidx}{-}{-} \in \fvalCtx\\\\
-%	\tKind{\tvarCtx}{\fvalCtx}{\tauidx}{\kappaidx}
-%}{
-%	\tKindX{\ttype{Tycon}{\tauidx}}{\kTypeBlur}
-%}
-%
-%\inferrule[k-ty-e]{
-%	\tKindX{\tau}{\kTypeBlur}\\
-%	\fval{Tycon}{\kappaidx}{-}{-} \in \fvalCtx\\\\
-%	\tKind{\tvarCtxX{x}{\kappaidx}}{\fvalCtx}{\tau_1}{\kappa}\\
-%	\tKindX{\tau_2}{\kappa}
-%}{
-%	\tKind{\tvarCtx}{\fvalCtx}{\tfamcase{\tau}{Tycon}{x}{\tau_1}{\tau_2}}{\kappa}
-%}
-%~~~~~~
-%\inferrule[k-d-i]{
-%	\tKindX{\tau}{\kTypeBlur}\\\\
-%	\isItermX{\ibar}
-%}{
-%	\tKindX{\tden{\ibar}{\tau}}{\kDen}
-%}
-%
-%\inferrule[k-d-i-checked]{
-%	\tKind{\Delta}{\fvalCtx}{\tau}{\kTypeBlur}\\\\
-%	\isItermX{\ibar}
-%}{
-%	\tKind{\Delta}{\fvalCtx}{\tden{\ibar}{\tau}^\checkmark}{\kDen}
-%}
-%~~~~~~~
-%\inferrule[k-d-e]{
-%	\tKindX{\tau}{\kDen}
-%}{
-%	\tKindX{\ttypeof{\tau}}{\kTypeBlur}
-%}
-%%
-%%\inferrule[iterm-intro]{
-%%	\isIterm{\tvarCtx}{\emptyctx}{\fvalCtx}{\iota}
-%%}{
-%%	\tKindX{\titerm{\iota}}{\kITerm}
-%%}
-%
-%\inferrule[k-reptype-i]{
-%	\isItype{\tvarCtx}{\fvalCtx}{\sbar}
-%}{
-%	\tKindX{\titype{\sbar}}{\kIType}
-%}
-%\end{mathpar}
-%$\fbox{$\isItermX{\ibar}$}$
-%\begin{mathpar}
-%\inferrule[k-i-var]{ }{
-%	\isItermX{\evar{x}}
-%}
-%~~~~~~
-%\inferrule[k-i-fix]{
-%	\isItypeX{\sbar}\\
-%	\isItermX{\ibar}
-%}{
-%	\isItermX{\ifix{\evar{x}}{\sbar}{\ibar}}
-%}
-%~~~~~~
-%\inferrule[k-i-transof]{
-%	\tKindX{\tau}{\kDen}
-%}{
-%	\isItermX{\itransof{\tau}}
-%}
-%%\inferrule[i-lam-kinding]{
-%%	\isItypeX{\sbar}\\\\
-%%	\isItermX{\ibar}
-%%}{
-%%	\isItermX{\ilam{\evar{x}}{\sbar}{\ibar}}
-%%}
-%
-%\text{\color{gray} (omitted rules are trivially recursive, like \textsc{k-i-fix})}
-%%\inferrule[iterm-dereify-kinding]{
-%%	\tKindX{\tau}{\kITerm}
-%%}{
-%%	\isItermX{\iup{\tau}}
-%%}
-%%
-%\end{mathpar}
-%$\fbox{$\isItypeX{\sbar}$}$
-%\begin{mathpar}
-%%\inferrule[i-int-kinding]{ }{
-%%	\isItypeX{\dint}
-%%}
-%%
-%%\inferrule[i-prod-kinding]{
-%%	\isItypeX{\sigma_1}\\\\
-%%	\isItypeX{\sigma_2}
-%%}{
-%%	\isItypeX{\dpair{\sigma_1}{\sigma_2}}
-%%}
-%%
-%\inferrule[k-s-unquote]{
-%	\tKindX{\tau}{\kIType}
-%}{
-%	\isItypeX{\dup{\tau}}
-%}
-%
-%\inferrule[k-s-repof]{
-%	\tKindX{\tau}{\kTypeBlur}
-%}{
-%	\isItypeX{\trepof{\tau}}
-%}
-%
-%\text{\color{gray} (omitted forms are trivially recursive)}
-%\end{mathpar}
-%\caption{\small Kinding for type-level terms.
-The kinding context $\tvarCtx$ maps type variables to kinds.
-Checked derivates (introduced by rule \textsc{k-d-i-checked}, discussed in Sec.
-\ref{repcon}) should only be constructed by the semantics (this can be enforced by a parser, so we will assume it in the metatheory).}
-%\label{tlkind}
-%\vspace{-10px}
-%\end{figure} 
-%In most languages, types are formed by applying one of a collection of \emph{type constructors} to zero or more \emph{indices}.
-In @$\lambda$, the situation is notionally similar.
-User-defined type constructors can be declared at the top of a program (or lifted to the top, in practice) using \textsf{tycon}.
-Each constructor in the program must have a unique name, written e.g.
-\fvar{Nat}.\footnote{We assume naming conflicts can be avoided by some extrinsic mechanism.} A type constructor must also declare an \emph{index kind}, $\kappaidx$.
-A type is introduced by applying a type constructor to an index of this kind, written $\ttype{Tycon}{\tauidx}$.
-%
-%@$\lambda$ supports, and makes extensive use of, simply-kinded type-level computation.
-Specifically, type-level terms, $\tau$, themselves form a typed lambda calculus.
-The classifiers of type-level terms are called \emph{kinds}, $\kappa$, to distinguish them from  \emph{types}.
-The kinding rules are specified in Fig.
-\ref{kindprog} and Fig.
-\ref{tlkind}.
-Types are  type-level values of kind $\kTypeBlur$, introduced as just described.
-The kind $\kTypeBlur$ also has an elimination form, $\tfamcase{\tau}{Tycon}{x}{\tau_1}{\tau_2}$, allowing the extraction of a type index by case analysis against a contextually-available type constructor.
-To a first approximation, one might think of type constructors as constructors of a built-in open datatype, $\kTypeBlur$, at the type-level.
-Like open datatypes, there is no notion of exhaustiveness so the default case is required for totality.
-%%We will write the kind of types as $\kTypeBlur$, though it is also written $\star$ or \verb|Type| in various similarly structured languages (see Sec. \ref{related-work}). 
-%% Rather than there being a fixed set of type constructors, we allow the programmer to declare new type  constructors, and give the static and dynamic semantics of their associated operators, by writing type-level functions.
-In the semantics for this calculus, our kind system combined with techniques borrowed from the typed compilation literature and a form of type abstraction allow us to prove strong type safety, decidability and conservativity theorems.
-%
-%
-% To permit the implementation of interesting type systems, the type-level language includes several kinds other than $\kTypeBlur$.
-We lift several  standard functional fragments to the type level: unit ($\kunit$), binary products ($\kpair{\kappa_1}{\kappa_2}$), binary sums ($\ksum{\kappa_1}{\kappa_2}$), lists ($\klist{\kappa}$) and integers ($\dint$).
-We also include labels ($\klabel$), written in a slanted font, e.g.
-$\tlabel{myLabel}$, which are string-like values that only support comparison and play a distinguished role in the expanded syntax, as we will later discuss.
-Our first example, $\fvar{Nat}$, is indexed trivially, i.e.
-by unit kind, $\kunit$, so there is only one natural number type, $\ttype{Nat}{\tunit}$, but we will show examples of type constructors that are indexed in more interesting ways in later portions of this work.
-For example, $\fvar{LabeledTuple}$ has index kind $\klist{\kpair{\klabel}{\kTypeBlur}}$.
-The type constructor $\fvar{Arrow}$ is included in the initial constructor context, $\fvalCtx_0$, shown in Fig.
-\ref{arrow}, and has index kind $\kpair{\kTypeBlur}{\kTypeBlur}$.
-% 
-% Type constructors are not first-class; they do not themselves have arrow kind as in some kind systems (e.g.
-\cite{watkins2008specifying}; Ch.
-22 of \emph{PFPL} describes a related system \cite{pfpl}).
-The type-level language does, however, include total functions of conventional arrow kind, written $\karrow{\kappa_1}{\kappa_2}$.
-Type constructor application can be wrapped in a type-level function to emulate a first-class or uncurried version of a type constructor for convenience (indeed, such a wrapper could be generated automatically, though we do not do so).
-%
-%Two type-level terms of kind $\kTypeBlur$ are equivalent if they apply the same constructor, identified by name, to equivalent indices.
-Going further, we ensure that deciding type equivalence requires only checking for syntactic equality after normalization by imposing the restriction that equivalence at a type constructor's index kind must be decidable in this way.
-Our treatment of equivalence in the type-level language is thus quite similar to the treatment of term-level equality using ``equality types'' in a language like Standard ML.
-A kind $\kappa$ is an  \emph{equality kind} if $\kEq{\kappa}$ can be derived (Fig.
-\ref{kindprog}).
-Conditional branching on the basis of equality at an equality kind can be performed in the type-level language (e.g.
-in \tvar{ifeq}, Fig.
-\ref{helpers}, discussed below).
-Equivalence at arrow kind is not decidable by our criteria, so type-level functions cannot appear within type indices.
-This also prevents general recursion from arising at the type level.
-Without this restriction, a type-level function taking a type as an argument could ``smuggle in'' a self reference as a type index, extracting it via case analysis (continuing our analogy to open datatypes, this is closely related to the positivity condition for inductive datatypes in total functional languages like Coq).% as maintaining the metatheoretic guarantee that typing respects type equivalence would impose a substantial burden in such a setting.% (a na\"ive approach to this would impose non-trivial extrinsic proof obligations onto extension developers that, unlike in others in this thesis, could threaten type safety).
-%
-%Every type constructor also defines a \emph{schema}, a type-level function that associates with every type an internal \emph{representation type}.
-We will return to this in Sec.
-\ref{repcon} after introducing operators.
-%\subsection{Operators and Operator Constructors}\label{opcons}
-%\begin{figure}[t]
-%\small
-%\[
-%\begin{array}{rcl}
-%\tvar{failure} & := & \tinr{\kDen}{\tunit}\\
-%\tvar{ok} & := & (\tlam{d}{\kDen}{\tinl{\kunit}{\tvar{d}}})\\
-%\tvar{ifeq} & := & (\tlam{t1}{\kTypeBlur}{\tlam{t2}{\kTypeBlur}{\tlam{d}{\kDen}{
-%	\\& & \quad \tifeq{\tvar{t1}}{\tvar{t2}}{\kTypeBlur}{\tvar{ok}~\tvar{d}}{\tvar{failure}}\\
-%\tvar{arity0} & := & (\tlam{a}{\klist{\kDen}}{\tlam{d}{(\ksum{\kDen}{\kunit})}{\tfold{\tvar{a}}{\tvar{d}}{\_}{\_}{\_}{\tvar{failure}}}})\\
-%\tvar{arity1} & := & (\tlam{a}{\klist{\kDen}}{\tlam{k}{\karrow{\kDen}{(\ksum{\kDen}{\kunit})}}{
-%	\\ & & \quad \tfold{\tvar{a}}{\tvar{tyerr}}{h}{t}{\_}{
-% \tvar{arity0}~\tvar{t}~(\tvar{k}~\tvar{h})}}})\\
-%\tvar{arity2} & := & (\tlam{a}{\klist{\kDen}}{\tlam{k}{\karrow{\kDen}{\karrow{\kDen}{(\ksum{\kDen}{\kunit})}}}{
-%	\\ & & \quad \tfold{\tvar{a}}{\tvar{tyerr}}{h}{t}{\_}{
-% \tvar{arity1}~\tvar{t}~(\tvar{k}~\tvar{h})}}})\\
-% \tvar{arity3} & := & (\tlam{a}{\klist{\kDen}}{\tlam{k}{\karrow{\kDen}{\karrow{\kDen}{\karrow{\kDen}{(\ksum{\kDen}{\kunit})}}}}{
-%	\\ & & \quad \tfold{\tvar{a}}{\tvar{tyerr}}{h}{t}{\_}{
-% \tvar{arity2}~\tvar{t}~(\tvar{k}~\tvar{h})}}})
-%}}}
-%\end{array}
-%\]
-%\caption{\small Useful (type-level) functions for writing operator definitions.}
-%\vspace{-10pt}
-%\label{helpers}
-%\end{figure}
-%\begin{figure}[t]
-%\small
-%\[
-%\begin{array}{rcl}
-%\Phi_0 & := & \emptyset, \fval{Arrow}{\kpair{\kTypeBlur}{\kTypeBlur}}{\tau_0}{\theta_0}\\
-%\tau_0 & := & \tlam{i}{\kpair{\kTypeBlur}{\kTypeBlur}}{\titype{\darrow{\trepof{\tfst{\tvar{i}}}}{\trepof{\tsnd{\tvar{i}}}}}}\\
-%\theta_0 & := & \tops{ap}{\kunit}{i}{a}{
-%	\tlam{i}{\kunit}{\tlam{a}{\klist{\kDen}}{
-%		\tvar{arity2}~\tvar{a}~\tlam{d1}{\kDen}{\tlam{d2}{\kDen}{
-%		\\& & \quad \tfamcase{\ttypeof{\tvar{d1}}}{Arrow}{x}{
-%		\\& & \quad\quad \tvar{ifeq}~\tfst{\tvar{x}}~\ttypeof{\tvar{d2}}
-%		\\& & \quad\quad\quad \tden{\itransof{\tvar{d1}}~\itransof{\tvar{d2}}}{\tsnd{\tvar{x}}}
-%		\\& & \quad\hspace{-3px}}{\tvar{tyerr}}
-%		}}
-%	}}
-%}
-%\end{array}
-%\]
-%\caption{\small The initial constructor context, $\fvalCtx_0$, defines the $\fvar{Arrow}$ type constructor.
-The introductory form, $\elam{x}{\tau}{e}$, is built into the language but the elimination form goes through the operator constructor $\fvar{Arrow}.\opvar{ap}$.}
-%\label{arrow}
-%\end{figure}
-%User-defined operator constructors are declared using \textsf{opcon}.
-For reasons that we will discuss, our calculus associates every operator  constructor with a type constructor.
-The \emph{fully-qualified name} of every operator constructor, e.g.
-$\fvar{Nat}.\opvar{z}$, must be unique.
-Operator constructors, like type constructors, declare an index kind, $\kappaidx$.
-In our first example, all the operator constructors are indexed trivially (by index kind $\kunit$), but other examples use more interesting indices.
-For example, $\fvar{LabeledTuple}.\opvar{prj}$ is the operator constructor used to access a field of a labeled tuple, so it has index kind $\klabel$.
-An operator itself is, notionally, selected by indexing an operator constructor, e.g.
-$\fvar{Nat}.\opvar{s}\langle \tunit \rangle$, but technically neither operator constructors nor operators are first-class at any level (additional machinery would be needed, e.g.
-an \textsf{Op} kind, but this is not fundamental to our calculus).
-%Instead, in the external language, an operator constructor is applied by simultaneously providing an index and  $n \geq 0$ \emph{arguments}, written $\eop{Tycon}{op}{\tauidx}{\splat{e}{1}{n}}$\footnote{It may be helpful to distinguish between type/operator constructors and \emph{term formers}, which are the metatheoretic entities that ultimately define the syntax.
-There are term formers at all levels in the calculus.
-For example, operator constructor application and $\lambda$ are  external term formers, and type constructor application is a type-level term former.
-An abstract syntax following Harper's conventions highlights this distinction \cite{pfpl}: the term formers $\mathtt{lam}$, $\mathtt{tcapp}$ and $\mathtt{ocapp}$, define, respectively the forms $\mathtt{lam}[\tau](x.e)$, $\texttt{tcapp}[\fvar{Tycon}](\tau)$ and $\mathtt{ocapp}[\fvar{Tycon}, \opvar{op}, \tauidx](\splat{e}{1}{n})$.}.
-For example, on line 18 of Fig.
-\ref{nat}, we see the operator constructors $\fvar{Nat}.\opvar{z}$ and $\fvar{Nat}.\opvar{s}$ being applied to compute $two$.
-In SML, the projection operator \verb|#3| can be applied to an $n$-tuple, $e$, iff $n \geq 3$.
-Note that it thus cannot be a function with a standard arrow type.
-Notionally, \verb|#| is an operator constructor and \verb|3| is its index.
-In an active embedding of $n$-tuples into @$\lambda$, this would be written $\eop{Tuple}{prj}{3}{e}$ (we will nearly recover ML's syntax later).
-%\footnote{Although our focus here is entirely on semantics, a brief note on syntax: in the expanded syntax, the trivial indices and empty argument lists can be omitted, so we could write \texttt{Nat.s(Nat.s(Nat.z))}.
-With the ability to ``open'' a type's operators into the context, we could shorten this still to \texttt{s(s(z))}.
-Alternatively, with the ability to define a TSL in a manner similar to that in Sec.
-\ref{aparsing}, we might instead just write \texttt{2}.}
-%
-%\subsection{Operator Definitions and Derivates}
-%The expressive and metatheoretic power of the calculus arises from how the rules for the active typing judgement handle this  operator constructor application form.
-Rather than fixing the specification of a finite collection of operator constructors and tasking the \emph{compiler} with deciding a typing derivation on its basis, the specification instead tasks the \emph{operator constructor's definition}: {a type-level function that must decide the type assignment and the translation on the basis of the index and arguments, or decide that this is not possible}.
-As indicated by rule \textsc{k-opcon} in Fig.
-\ref{kindprog}, an operator constructor with index kind $\kappaidx$ must have a definition of kind $$\kappaidx \rightarrow \klist{\kDen} \rightarrow (\kDen + \kunit)$$ As this suggests, the active typing judgement  will provide the operator index and a list of {recursively determined \emph{derivates}}, which have kind $\kDen$, for the arguments and ask the definition to return a value of ``option kind'', $\ksum{\kDen}{\kunit}$, where the trivial case indicates that a derivate cannot be constructed due to an invalid index, an incorrect number of arguments or an argument of an invalid type.\footnote{In practice, we would require operator constructor providers to report information about the precise location of the error (e.g.
-which argument was of incorrect type) and provide an appropriate error message and other metadata, but we omit this in the semantics.}
-%
-%A \emph{derivate} is introduced in operator definitions by the form $\tden{\ibar}{\tau}$, where $\tau$ is a type assignment and has kind $\kTypeBlur$ and $\ibar$ is a \emph{translational internal term}.
-The syntax for translational internal terms mirrors that of internal terms, $\iota$, adding an additional form, $\itransof{\tauD}$, that permits one translation to be composed using the translations of other derivates (e.g.
-the derivate of an argument).
-Fig.
-\ref{tlkind} shows the kinding rules for derivates and translational internal terms (we will discuss checked derivates shortly; they are only used internally by the semantics).
-The type assignment can be extracted from a derivate using $\ttypeof{\tauD}$.
-%
-%Let us return to Fig. \ref{nat} to build intuition. 
-%The definition of $\fvar{Nat}.\opvar{z}$ in Fig.
-\ref{nat} is quite simple: it returns the derivate $\tden{{0}}{\ttype{Nat}{\tunit}}$ if no arguments were provided, and indicates an error otherwise.
-This is done by calling a simple helper function, $\tvar{arity0}$, seen in Fig.
-\ref{helpers}.
-The definition of $\fvar{Nat}.\opvar{s}$ is only slightly more complex, because it requires inspecting a single argument.
-The helper function $\tvar{arity1}$ checks the arity of the argument list.
-If it is correct, it passes the single derivate to the shown continuation, returning the error case otherwise (which we give a more descriptive name, $\tvar{failure}$).
-The continuation we provided checks that the type of the argument is $\ttype{Nat}{\tunit}$ using a similar helper function, $\tvar{ifeq}$.
-If so, it constructs the derivate $\tden{\itransof{\tvar{d}}+1}{\ttype{Nat}{\tunit}}$ and $\tvar{ifeq}$ wraps it with $\tvar{ok}$.
-We will return to the recursor shortly.
-%
-%These two operator definitions make the following rules admissible in a constructor context, let us call it $\fvalCtx_{\mathbf{T}}$, containing the initial constructor context (Fig.
-\ref{arrow}) and the constructors of  Fig.
-\ref{nat} and any typing context well-kinded under that constructor context (the rules governing context kinding are in the appendix):
-%\newcommand{\fvalT}{\fvalCtx_{\mathbf{T}}}
-%\begin{mathpar}
-%\inferrule[gt-z]{ }{
-%	\ecompiles{\Upsilon}{\fvalT}{\eop{Nat}{z}{\tunit}{}}{\ttype{Nat}{\tunit}}{0}
-%}
-%
-%\inferrule[gt-s]{
-%	\ecompiles{\Upsilon}{\fvalT}{e}{\ttype{Nat}{\tunit}}{\iota}
-%}{
-%	\ecompiles{\Upsilon}{\fvalT}{\eop{Nat}{s}{\tunit}{e}}{\ttype{Nat}{\tunit}}{\iota + 1}
-%}
-%\end{mathpar}
-%
-% We will consider the complexities shortly, but it requires little more than basic intuition to see that we can strongly embed the types and introductory forms of G\"odel's $\mathbf{T}$ in @$\lambda$ quite directly (adding essentially only a barrage of trivial indices which will be removed syntactically in the expanded syntax).
-If we keep the constructor context fixed, we should be able to prove the embedding correct inductively.
-We will define this embedding later\todo{refer to it now}.
-%
-% Figure \ref{att} shows how $\fvalCtx_{\mathbf{T}}$ would be built up and how these conclusions could be derived.
-The active typing judgement simply defers to the \emph{checked active typing judgement} (Fig.
-\ref{absatt}) to first produce a type assignment and a translational internal term, called the \emph{abstract translation}.
-It then \emph{deabstracts} the abstract translation to produce a type assignment and translation (Fig.
-\ref{deabstraction}).
-% 
-% The rule \textsc{att-op} shows how operator constructor definitions are invoked.
-First, type assignments  and translational internal terms are recursively derived.
-Then, the operator constructor definition, $\taudef$ is extracted from the constructor context and applied to the index, $\tauidx$, and a list of \emph{checked derivates} constructed from the derivations produced in the first step.
-If this produces a derivate, $\tauD$, via the left case of the sum, then it is itself, finally, \emph{checked} to produce a type assignment, $\tau$, and translational internal term, $\ibar$.
-% 
-%The normalization rules for type-level terms are shown in Fig. \ref{tleval} and are rather uninteresting. We will discuss their metatheoretic properties in Sec. \ref{theory}.
-% 
-% Derivate checking, shown in Fig. \ref{ait}, is described in the next two subsections. 
-% Note that derivate checks are still static checks; they do not change the translation.
-Derivate checking is sufficient to guarantee type safety and preclude interfere issues between extensions, as we will describe, because taking a path through an operator constructor definition that produces a problematic derivate will fail.
-If providers do not want to burden clients with this variety of error, which would require them to debug the constructor definition itself, we will discuss how constructor correctness could be verified mechanically by strengthening the kind system in Sec.
-\ref{limitations}.
-%We will discuss how to prove that an extension  operator definition could be \emph{proven} to not produce such derivates (so that users only ever face type errors arising from their own code, not also extension correctness issues) in Sec.
-\ref{limitations}.
-The checks are sophisticated enough that they cannot be eliminated by a simple kind system.
-%
-%%$\tvar{pop\_final} : \karrow{\karrow{\klist{\kDen}}{(\karrow{\kTypeBlur}{\karrow{\kITerm}{(\ksum{\kDen}{\kunit})})}}}{(\ksum{\kDen}{\kunit})}$ ``pops'' a derivate from the head of the list and, if no other arguments remain, passes its translation and type to the ``continuation'', returning the error case otherwise.
-The continuation checks if the argument type is equal to $\ttype{Nat}{\tunit}$ using $\tvar{check\_type} : \karrow{\kTypeBlur}{\karrow{\kTypeBlur}{\karrow{\kDen}{(\ksum{\kDen}{\kunit})}}}$, which operates similarly.
-If these arity and type checks succeed, the resulting derivate is composed by adding one to the translation of the argument, passed into the continuation as $\tvar{x}$ in our example, and pairing it with the natural number type: $\tden{\titerm{\iup{\tvar{x}} + 1}}{\ttype{Nat}{\tunit}}$.
-We will return to the definition of $\fvar{Nat}.\opvar{rec}$ after in the next subsection.
-%
-%%We leave adding the ability to declare and manipulate new contexts as future work.
-% 
-%\subsection{Representational Consistency Implies Type Safety}\label{repcon}
-%\begin{figure}[t]
-%\small
-%$\fbox{\inferrule{}{\pcompilesX{\rho}}}$
-%\begin{mathpar}
-%\inferrule[att-tycon]{
-%	\pcompiles{\fvalCtxX{\fvalDf}}{\rho}{\iota}
-%}{
-%	\pcompilesX{\pfam{\familyDf}{\rho}}
-%}
-%%
-%%\inferrule[att-def]{
-%%	\tEvalX{\tau}{\tau'}\\
-%%	\pcompilesX{\subst{\tau'}{\tvar{t}}{\rho}}
-%%}{
-%%	\pcompilesX{\pdef{t}{\kappa}{\tau}{\rho}}
-%%}
-%
-%\inferrule[att-e-prog]{
-%	\ecompiles{\emptyctx}{\fvalCtx}{e}{\tau}{\iota}%\\
-%%	\eraseX{\gabs}{\iota}
-%}{
-%	\pcompilesX{e}
-%}
-%\end{mathpar}
-%$\fbox{\inferrule{}{\ecompilesX{e}{\tau}{\iota}}}$
-%~~~$\etCtx ::= \emptyctx \pipe \etCtxX{x}{\tau}$
-%\begin{mathpar}
-%\inferrule[att-e]{
-%	\ecompilesA{\Upsilon}{\fvalCtx}{e}{\tau}{\ibar}\\
-%	\eraseX{\ibar}{\iota}
-%}{
-%	\ecompiles{\Upsilon}{\fvalCtx}{e}{\tau}{\iota}\\
-%}
-%\end{mathpar}
-%\caption{\small Active typing judgement for programs and external terms. The deabstraction ($\lightning$) judgement is in Fig. \ref{deabstraction}.}
-%\label{att}
-%%\vspace{-10pt}
-%\end{figure}
-%\begin{figure}[t]
-%\small
-%$\fbox{\inferrule{}{\ecompilesAX{e}{\tau}{\ibar}}}$
-%\begin{mathpar}
-%\inferrule[att-var]{ }{
-%	\ecompilesA{\etCtxX{x}{\tau}}{\fvalCtx}{\evar{x}}{\tau}{\evar{x}}
-%}
-%
-%\inferrule[att-lam]{
-%	\tEvalX{\tau_1}{\tau_1'}\\
-%%	\ddbar{\fvar{Arrow}}{\fvalCtx}{\trepof{\tau_1'}}{\sbar_1}\\
-%	%\delfromtau{$\Xi_0$}{\fvalCtx}{\tau_1'}{\sabs}\\\\
-%	\ecompilesA{\etCtxX{x}{\tau_1'}}{\fvalCtx}{e}{\tau_2}{\ibar}
-%}{
-%	\ecompilesA{\etCtx}{\fvalCtx}{\elam{\evar{x}}{\tau_1}{e}}{\ttype{Arrow}{\tpair{\tau_1'}{\tau_2}}}{\ilam{\evar{x}}{\trepof{\tau_1'}}{\ibar}}
-%}
-%
-%\inferrule[att-op]{
-%%	\tEvalX{\tauidx}{\tau_{\text{idx}}'}\\\\
-%	\ecompilesA{\etCtx}{\fvalCtx}{e_1}{\tau_1}{\ibar_1}~~~~
-%	\cdots~~~~
-%	\ecompilesA{\etCtx}{\fvalCtx}{e_n}{\tau_n}{\ibar_n}\\
-%	\fval{Tycon}{-}{-}{\theta} \in \fvalCtx\\
-%	\tops{op}{\kappaidx}{i}{a}{\taudef} \in \theta\\
-%	\tEvalX{\taudef~\tauidx~(\tden{\ibar_1}{\tau_1}^\checkmark :: \cdots :: \tden{\ibar_n}{\tau_n}^\checkmark :: \tnil{\kDen})}{\tinl{\kunit}{\tauD}}\\
-%	\dcheck{\Upsilon}{\fvalCtx}{\fvar{Arrow}, \fvar{Tycon}}{\tauD}{\tau}{\ibar}
-%}{
-%	\ecompilesA{\etCtx}{\fvalCtx}{\eop{Tycon}{op}{\tauidx}{\splat{e}{1}{n}}}{\tau}{\ibar}
-%}
-%\end{mathpar}
-%\caption{\small The checked active typing judgement.
-The normalization judgement for type-level terms ($\Downarrow$) is specified in Fig.
-\ref{tleval}.
-Derivate checking is specified in Fig.
-\ref{ait}.
-Note that checked derivates can \emph{only} be constructed by the \textsc{att-op} rule (we enforce this judgementally in Sec.
-\ref{theory}).}
-%\label{absatt}
-%\end{figure}
-%In our example, $\fvar{Nat}.\opvar{z}$ produces a translation of internal type $\dint$.
-For the translation produced by $\fvar{Nat}.\opvar{s}$ to be well-typed requires that $\itransof{\tvar{d1}}$ be of internal type $\dint$.
-Because we have checked that $\ttypeof{\tvar{d1}}$ is $\ttype{Nat}{\tunit}$ and the only other introductory form is $\fvar{Nat}.\opvar{z}$, this is not an issue.
-Well-typed external terms of type $\ttype{Nat}{\tunit}$ will always translate to well-typed internal terms of internal type $\dint$ as given.
-%
-%This would not be true if, for example, we added an operator constructor $\fvar{Nat}.\opvar{z2}$ that produced the derivate $$\tden{{(0, ())}}{\ttype{Nat}{\tunit}}$$
-%\noindent
-%In this case, there would be two different internal types, $\dint$ and $\dpair{\dint}{\dunit}$,  associated with a single external type, $\ttype{Nat}{\tunit}$.
-This makes it impossible to reason \emph{compositionally} about the translation of an external term of type $\ttype{Nat}{\tunit}$, so our implementation of $\fvar{Nat}.\opvar{s}$ would produce ill-typed translations in some cases but not others.
-Similarly, we wouldn't be able to write functions over all natural numbers because there would not be a well-typed translation to give to such a function.
-This violates type safety: there are now well-typed external terms, according to the active typing judgement, for which evaluating the corresponding translation would ``go wrong''.
-%
-%To reason compositionally about the semantics of well-typed external terms when they are given meaning by translation to a typed internal language, we must have that the statics maintains the following property: for every  type, $\tau$, there must exist an internal type, $\sigma$, called its \emph{representation type}, such that the translation of every external term of type $\tau$ has internal type $\sigma$.
-This principle of \emph{representational consistency} arises essentially as a strengthening of the inductive hypothesis necessary to prove that all well-typed external terms translate to well-typed internal terms, precisely because $\lambda$ and operators like $\fvar{Nat}.\opvar{s}$ are defined compositionally.
-It is closely related to the concept of \emph{type-preserving compilation} developed by Morrisett et al.
-for the TIL compiler for SML \cite{tarditi+:til-OLD}.
-%That is, instead of being used as a necessary condition for compiler correctness, we are using it as a sufficient condition for type safety.
-%%We emphasize the distinction between translation (which in our calculus endows terms with a dynamic semantics) and compilation (which must preserve the dynamic semantics of terms).
-%In the presence of extensions, checking for representational consistency becomes subtle if we wish to guarantee conservativity, as we will discuss in the next subsection.
-%
-%For the semantics to ensure that representational consistency is maintained by all operator definitions, we require, as briefly mentioned in Sec.
-\ref{types}, that each type constructor must declare a \emph{representation schema} after the keyword \textsf{schema}.
-This must be a type-level function of kind $\karrow{\kappaidx}{\kIType}$, where $\kappaidx$ is the index kind of the type constructor.
-When the semantics (i.e.
-the compiler) needs to determine the representation type of the type $\ttype{Tycon}{\tauidx}$ it simply applies the representation schema of $\fvar{Tycon}$ to ${\tauidx}$.
-%
-%The kind $\kIType$ has introductory form $\titype{\bar \sigma}$ and no elimination form.
-It is not an equality kind.
-There are two forms of \emph{translational internal type}, $\bar \sigma$, that do not correspond to forms in $\sigma$ that, as with translational internal terms, allow  compositional formation:
-%\begin{enumerate}
-%\item $\dup{\tau}$ ``unquotes'' the translational internal type $\tau$, so that internal types can be formed programmatically\footnote{An analagous form could be added to translational internal terms, but the same effect can be achieved by working directly with derivates, albeit with some inconvenience when ``temporary intermediates'' are needed.} \todo{example}
-%\item $\trepof{\tau}$ refers to the representation type of type $\tau$. This type may have come from an argument derivate or extracted from a type or operator index.
-%\end{enumerate}
-%
-%These additional forms are not needed by the representation schema of $\fvar{Nat}$  because it is trivially indexed.
-In Fig.
-\ref{tuple}, we show an example of the type constructor $\fvar{Tuple}$, implementing the semantics of $n$-tuples by translation to nested binary products.
-Here, the representation schema requires referring to the representations of the tuple's constituent types, given in the type index\todo{add this example from appendix of ESOP}.
-%
-%Translational internal terms also contain translational internal types.
-We see this in the definition of the recursor on natural numbers, $\fvar{Nat}.\opvar{rec}$.
-After checking the arity, extracting the derivates of its three arguments, and checking their types (using \fvar{Arrow} to bind variables in the third position), it produces a derivate that implements the necessarily recursive dynamic semantics using a fixpoint computation in the internal language.
-This derivate's type assignment is the arbitrary type, $\tvar{t2}$.
-We cannot know what the representation type of $\tvar{t2}$ is (even in a closed constructor context, because the schema is not directly callable), so we refer to it abstractly using $\trepof{\tvar{t2}}$.
-Luckily, the translation is well-typed \emph{independent of the representation type}.
-In other words, a proof of representational consistency of the derivate produced by the definition of $\fvar{Nat}.\opvar{rec}$ is parametric (in a metamathematical sense here) over the representation type of $\tvar{t2}$.
-In fact, we will demand this for all types constructed by a type constructor other than the one an operator constructor is associated with.
-This leads us into the concept of conservativity.
-%
-%
-%\subsection{Representational Independence Implies Conservativity}\label{repind}
-%In isolation (let us call this language @$\lambda[\fvalCtx_{\mathbf{T}}]$) our definition of natural numbers can be shown to be a strong encoding of the semantics of G\"odel's $\mathbf{T}$.
-That is, there is a mapping of terms and types into @$\lambda[\fvalCtx_{\mathbf{T}}]$ and an inverse mapping.
-Indeed, these are the entirely obvious one, given how directly we constructed @$\lambda[\fvalCtx_{\mathbf{T}}]$.
-%
-%Typing is preserved by  these mappings, and we can show by a bisimulation argument that the translation produced by @$\lambda[\fvalCtx_{\mathbf{T}}]$ implements the dynamic semantics of G\"odel's $\mathbf{T}$.
-We must leave full details to an appendix, but the techniques are fundamentally standard when reasoning about compiler correctness, here lifted into the language.
-A necessary lemma in the proofs, however, is that the value of every translation of an external term of type $\ttype{Nat}{\tunit}$ is a non-negative integer.
-This is needed to show that the fixpoint computation in the definition of $\fvar{Nat}.\opvar{rec}$ is terminating, as the recursor always does in $\mathbf{T}$.
-A very similar lemma is needed to show that the translation of every external term of type $\ttype{Nat}{\tunit}$ is equivalent to the translation that would be produced by some combination of applications of $\fvar{Nat}.\opvar{z}$ and $\fvar{Nat}.\opvar{s}$ (corresponding to the canonical forms lemma of $\mathbf{T}$).
-%
-%Fortunately, the definitions we have provided thus far admit this lemma.
-Indeed, the proof is quite straightforward.
-However, it is also quite precarious because it relies on exhaustive induction over the available operators.
-As soon as we load another operator or type constructor into the language, these theorems may no longer be \emph{conserved}.
-We already saw an example where a representationally inconsistent operator caused problems, but the semantics render uses of it ill-typed using the representation schema (in a manner we will make more precise shortly).
-But there are operator constructors that are representationally consistent, but still cause problems because they violate internal invariants.
-For example, $$\tops{n1}{\kunit}{i}{a}{\tden{-1}{\ttype{Nat}{\tunit}}}$$
-%
-%Were this operator constructor introduced inside $\fvar{Nat}$, there is no way for our semantics to distinguish between an incorrect implementation of $\fvar{Nat}$ and a correct implementation of some other type where non-termination of the recursor was the intended behavior (without a separate specification).
-%
-%But in any \emph{other} type constructor, this clearly appears to violate the tacit, thus far, abstraction barrier between type constructors.
-In particular, if we extended @$\lambda[\fvalCtx_{\mathbf{T}}]$ with such a type constructor, let us call it $\fvar{Evil}$, then our crucial lemma relating to non-negativity would not be conserved, and thus, our carefully constructed strong encoding would no longer be a strong encoding.
-In particular, our canonical forms lemma now has a new case, and the recursor behaves badly.
-\emph{If we assume an abstraction barrier between type constructors\footnote{A fragment consisting of finitely many type constructors can be easily turned into just one by adding a sum type to the index and case analyzing against it whenever a distinction is necessary.} must exist, we do not need need a specification to see that $\fvar{Evil}.\opvar{n1}$ is violating it}.
-We can prevent problems by using a technique closely related to the one used to enforce abstraction barriers between modules -- holding the representation type abstract.
-If $\fvar{Evil}.\opvar{n1}$ cannot know that the representation type of $\ttype{Nat}{\tunit}$ is $\dint$, then its definition is incorrect independent of the specification of $\fvar{Nat}$.
-%\section{Metatheory}\label{theory}
-%\subsection{Active Typing}
-%\begin{figure}[t]
-%\small
-%\noindent
-%$\fbox{\inferrule{}{\dcheckX{\tauD}{\tau}{\ibar}}}$
-%~~~$\Xi ::= \emptyset \pipe \Xi, \fvar{Tycon}$
-%\begin{mathpar}
-%\inferrule[d-check]{
-%	\ddbar{\Xi}{\fvalCtx}{\trepof{\tau}}{\sabs}\\
-%%	\eCtxTogCtx{\Xi}{\fvalCtx}{\etCtx}{\gtCtx}\\
-%	\checkRC{\emptyctx}{\Xi}{\fvalCtx}{\ibar}{\sabs}
-%}{
-%	\dcheckX{\tden{\ibar}{\tau}}{\tau}{\ibar}
-%}
-%
-%\inferrule[d-checked]{
-%	%\dcheckX{\tden{\ibar}{\tau}}{\tau}{\ibar}
-%}{
-%	\dcheckX{\tden{\ibar}{\tau}^\checkmark}{\tau}{\ibar}
-%}
-%\end{mathpar}
-%$\fbox{\inferrule{}{\ddbarX{\sbar}{\sabs}}}$
-%~~~$\sabs ::= \sabs \rightharpoonup \sabs' \pipe \cdots \pipe \sabsrep{\tau}$
-%%~~~$\fvalCtx ::= \fvalCtx_0 \pipe \fvalCtx, \fvalDf$~~~
-%\begin{mathpar}
-%%\inferrule[pconc-int]{ }{
-%%	\ddbarX{\dint}{\dint}
-%%}
-%\inferrule[abs-trity-parr]{
-%	\ddbarX{\sbar_1}{\sabs_1}\\
-%	\ddbarX{\sbar_2}{\sabs_2}
-%}{
-%	\ddbarX{\darrow{\sbar_1}{\sbar_2}}{\darrow{\sabs_1}{\sabs_2}}
-%}
-%
-%\inferrule[abs-trity-cancel]{
-%	\ddbarX{\sbar}{\sabs}
-%}{
-%	\ddbarX{\dup{\titype{\sbar}}}{\sabs}
-%}
-%
-%\inferrule[abs-trity-concrete]{
-%	\fvar{Tycon} \in \Xi\\
-%	\fval{Tycon}{-}{\taurep}{-} \in \fvalCtx\\\\
-%	\tEvalX{\taurep~\tauidx}{\titype{\sbar}}\\
-%	\ddbarX{\sbar}{\sabs}
-%}{
-%	\ddbar{\Xi}{\fvalCtx}{\trepof{\ttype{Tycon}{\tauidx}}}{\sabs}
-%}
-%
-%\inferrule[abs-trity-abstract]{
-%	\fvar{Tycon} \notin \Xi
-%}{
-%	\ddbarX{\trepof{\ttype{Tycon}{\tauidx}}}{\sabsrep{\ttype{Tycon}{\tauidx}}}
-%}
-%%
-%%\inferrule[pconc-prod]{
-%%	\ddbarX{\sbar_1}{\sbar_1'}\\
-%%	\ddbarX{\sbar_2}{\sbar_2'}
-%%}{
-%%	\ddbarX{\dpair{\sbar_1}{\sbar_2}}{\dpair{\sbar_1'}{\sbar_2'}}
-%%}
-%%
-%
-%\text{\color{gray} (omitted rules follow recursively like \textsc{abs-it-parr})}
-%\end{mathpar}
-%%$\fbox{\inferrule{}{\eCtxTogCtxX{\etCtx}{\gtCtx}}}$
-%%~~~$\gtCtx ::= \emptyctx \pipe \gtCtxX{x}{\sbar}$
-%%\begin{mathpar}
-%%\inferrule[abs-empty]{ }{
-%%	\eCtxTogCtxX{\emptyctx}{\emptyctx}
-%%}
-%%
-%%\inferrule[abs-ctx]{
-%%	\eCtxTogCtxX{\etCtx}{\gtCtx}\\
-%%	\ddbarX{\trepof{\tau}}{\sbar}
-%%}{
-%%	\eCtxTogCtxX{\etCtxX{x}{\tau}}{\gtCtxX{x}{\sbar}}
-%%}
-%%\end{mathpar}
-%$\fbox{\inferrule{}{\checkRCX{\ibar}{\sabs}}}$
-%~~~$\gtCtx ::= \emptyset \pipe \gtCtx, x \sim \sbar$
-%\begin{mathpar}
-%\inferrule[abs-i-var]{ }{
-%	\checkRC{\gtCtxX{x}{\sbar}}{\Xi}{\fvalCtx}{\evar{x}}{\sbar}
-%}
-%
-%\inferrule[abs-i-fix]{
-%	\ddbarX{\sbar}{\sabs}\\
-%	\checkRC{\gtCtxX{x}{\sabs}}{\Xi}{\fvalCtx}{\ibar}{\sabs}
-%}{
-%	\checkRCX{\ifix{x}{\sbar}{\ibar}}{\sbar'}
-%}
-%%
-%%\inferrule[abs-i-lam]{
-%%	\ddbarX{\sbar_1}{\sbar_1'}\\\\
-%%	\checkRC{\gtCtxX{x}{\sbar_1'}}{\Xi}{\fvalCtx}{\ibar}{\sbar_2}
-%%}{
-%%	\checkRCX{\ilam{x}{\sbar_1}{\ibar}}{\darrow{\sbar_1'}{\sbar_2}}
-%%}
-%%
-%%\inferrule[abs-i-ap]{
-%%	\checkRCX{\ibar_1}{\darrow{\sbar_1}{\sbar_2}}\\\\
-%%	\checkRCX{\ibar_2}{\sbar_1}
-%%}{
-%%	\checkRCX{\iapp{\ibar_1}{\ibar_2}}{\sbar_2}
-%%}
-%
-%\text{\color{gray} (omitted rules follow from a standard statics  similarly)}
-%
-%%\inferrule[abs int]{ }{
-%%	\checkRCX{\iintlit}{\dint}
-%%}
-%%
-%%\inferrule[abs op]{
-%%	\checkRCX{\ibar_1}{\dint}\\
-%%	\checkRCX{\ibar_2}{\dint}
-%%}{
-%%	\checkRCX{\iop{\ibar_1}{\ibar_2}}{\dint}
-%%}
-%%
-%%\inferrule[abs pair]{
-%%	\checkRCX{\ibar_1}{\sbar_1}\\
-%%	\checkRCX{\ibar_2}{\sbar_2}
-%%}{
-%%	\checkRCX{\ipair{\ibar_1}{\ibar_2}}{\dpair{\sbar_1}{\sbar_2}}
-%%}
-%%
-%%\inferrule[abs fst]{
-%%	\checkRCX{\ibar}{\dpair{\sbar_1}{\sbar_2}}
-%%}{
-%%	\checkRCX{\ifst{\ibar}}{\sbar_1}
-%%}
-%%
-%%\inferrule[abs snd]{
-%%	\checkRCX{\ibar}{\dpair{\sbar_1}{\sbar_2}}
-%%}{
-%%	\checkRCX{\isnd{\ibar}}{\sbar_2}
-%%}
-%%
-%% \inferrule[abs-int-eq]{
-%% 	\checkRCX{\ibar_1}{\dint}\\
-%% 	\checkRCX{\ibar_2}{\dint}\\\\
-%% 	\checkRCX{\ibar_3}{\sbar}\\
-%% 	\checkRCX{\ibar_4}{\sbar}
-%% }{
-%% 	\checkRCX{\iIfEq{\ibar_1}{\ibar_2}{\dint}{\ibar_3}{\ibar_4}}{\sbar}
-%% }
-%%\inferrule[abs-iterm-inverse]{
-%%	\checkRCX{\ibar}{\sbar}
-%%}{
-%%	\checkRCX{\iup{\titerm{\ibar}}}{\sbar}
-%%}
-%%
-%\inferrule[abs-trans-local-unchecked]{
-%	\fvar{Tycon} \in \Xi\\
-%	\ddbarX{\trepof{\ttype{Tycon}{\tauidx}}}{\sabs}\\
-%	\checkRCX{\ibar}{\sabs}
-%}{
-%	\checkRCX{\itransof{\tden{\ibar}{\ttype{Tycon}{\tauidx}}}}{\sabs}
-%}
-%
-%\inferrule[abs-trans-local-checked]{
-%	\fvar{Tycon} \in \Xi\\
-%	\ddbarX{\trepof{\ttype{Tycon}{\tauidx}}}{\sabs}
-%}{
-%	\checkRCX{\itransof{\tden{\ibar}{\ttype{Tycon}{\tauidx}}^\checkmark}}{\sabs}
-%}
-%
-%\inferrule[abs-trans-foreign-checked]{
-%	\fvar{Tycon} \notin \Xi
-%}{
-%	\checkRCX{\itransof{\tden{\ibar}{\ttype{Tycon}{\tauidx}}^{\checkmark}}}{\sabsrep{\ttype{Tycon}{\tauidx}}}
-%}
-%\end{mathpar}
-%\caption{\small Abstracted internal typing involves typechecking the internal language while holding the representation of all types other than those in the current fragment, listed in the fragment context, $\Xi$, abstract.
-They key rule is \textsc{abs-trans-foreign-checked}, permitting the translation of a checked derivate that has a type assignment outside the fragment to be used against that type's abstract representation type.}
-%\label{ait}
-%\end{figure}
-%\begin{figure}
-%\small
-%$\fbox{\inferrule{}{\eraseX{\sbar}{\sigma}}}$
-%\begin{mathpar}
-%\inferrule[deabs-it-repof]{
-%	\fval{Tycon}{-}{\taurep}{-} \in \fvalCtx\\
-%	\tEvalX{\taurep~\tauidx}{\titype{\sbar}}\\
-%	\eraseX{\sbar}{\sigma}
-%}{
-%	\eraseX{\trepof{\ttype{Tycon}{\tauidx}}}{\sigma}
-%}
-%
-%\inferrule[deabs-it-cancel]{
-%	\eraseX{\sbar}{\sigma}
-%}{
-%	\eraseX{\dup{\titype{\sbar}}}{\sigma}
-%}
-%
-%\inferrule[deabs-it-parr]{
-%	\eraseX{\sbar_1}{\sigma_1}\\
-%	\eraseX{\sbar_2}{\sigma_2}
-%}{
-%	\eraseX{\darrow{\sbar_1}{\sbar_2}}{\darrow{\sigma_1}{\sigma_2}}
-%}
-%
-%\text{\color{gray} (remaining forms follow recursively like \textsc{deabs-it-parr}, omitted)}
-%\end{mathpar}
-%$\fbox{\inferrule{}{\eraseX{\ibar}{\iota}}}$
-%\begin{mathpar}
-%\inferrule[deabs-i-var]{ }{\eraseX{x}{x}}
-%
-%\inferrule[deabs-i-fix]{
-%	\eraseX{\sbar}{\sigma}\\
-%	\eraseX{\ibar}{\iota}
-%}{
-%	\eraseX{\ifix{x}{\sbar}{\ibar}}{\ifix{x}{\sigma}{\iota}}
-%}
-%
-%\text{\color{gray} (remaining forms follow recursively like \textsc{deabs-i-fix}, omitted)}
-%
-%\inferrule[deabs-trans-unchecked]{
-%	\eraseX{\ibar}{\iota}
-%}{
-%	\eraseX{\itransof{\tden{\ibar}{\tau}}}{\iota}
-%}
-%
-%\inferrule[deabs-trans-checked]{
-%	\eraseX{\ibar}{\iota}
-%}{
-%	\eraseX{\itransof{\tden{\ibar}{\tau}^\checkmark}}{\iota}
-%}
-%\end{mathpar}
-%\caption{\small Deabstraction simply recursively eliminates syntactic abstraction barriers and composition-related constructs, and is the final step.}
-%\label{deabstraction}
-%\end{figure}
-%
-%allowing us to prove decidability of typechecking for the language as a whole by essentially just citing the termination theorem for the type-level language and proving that the underlying semantics are otherwise well-founded.
-%
-%\section{Expanded Syntax}\label{expanded-syntax}
-%
-%\section{Examples}\label{examples}
-%\begin{figure}
-%\small
-%\begin{flalign}
-%\small
-% & \family{Tuple}{\klist{\kTypeBlur}}{\\
-% & \quad \tops{new}{\kunit}{\_}{a}{\tlam{\_}{\kunit}{\tlam{a}{\klist{\kDen}}{\\
-% & \quad\quad
-% 	\tfold{\tvar{a}}{\tden{{()}}{\ttype{Tuple}{\tnil{\kTypeBlur}}}}{d}{b}{r}{\\
-%			& \quad\quad\quad \tfamcase{\ttypeof{\tvar{r}}}{Ntuple}{i}{\\
-%			& \quad\quad\quad\quad \tfold{\tvar{b}}{\tden{\itransof{\tvar{d}}}{\ttype{Ntuple}{\tcons{\ttypeof{\tvar{d}}}{\tvar{i}}}}\\& \quad\quad\quad\quad}{\_}{\_}{\_}{\tden{\\& \quad\quad\quad\quad\quad {(\itransof{\tvar{d}},\itransof{\tvar{r}})}}{\ttype{Ntuple}{\tcons{{\ttypeof{\tvar{d}}}}{\tvar{i}}}}}
-%			}{\terr}
-%	}
-% }}};\\
-% & \quad \tops{prj}{\kint}{i}{a}{\lambda \tvar{i}:\kint.\lambda \tvar{a}:\klist{\kDen}. 
-% 	\tapp{\tvar{arity1}}{\tapp{\tvar{a}}{\tlam{d}{\kDen}{\\
-%	& \quad\quad\tfamcase{\ttypeof{\tvar{d}}}{Ntuple}{nl}{\\
-% & \quad\quad \tfold{\tvar{nl}}{\terr}{t1}{j}{\_}{\\
-% & \quad\quad \tfold{\tvar{j}}{\tifeq{\tvar{i}}{1}{\dint}{\tden{\itransof{d}}{\tvar{t1}}}{\terr}\\&\quad\quad\quad}{\_}{\_}{\_}{\\
-% & \quad\quad\quad (\tlam{p}{\kpair{\kDen}{\dint}}{\tifeq{\tsnd{\tvar{p}}}{\tvar{i}}{\dint}{\tfst{\tvar{p}}}{\terr}})\\
-% & \quad\quad\quad (\tapp{\tvar{foldl}}{\tapp{\tvar{nl}}{\tapp{(\tden{\tvar{x}}{\tvar{nt}},0)}{
-% 	\tlam{r}{\kpair{\kDen}{\dint}}{\tlam{t}{\kTypeBlur}{\tlam{ts}{\klist{\kTypeBlur}}{\\
-%	& \quad\quad\quad\quad \tifeq{\tvar{i}}{\tsnd{\tvar{r}}}{\dint}{\tvar{r}}{\tdencase{\tfst{\tvar{r}}}{rx}{\_}{\\
-%	& \quad\quad\quad\quad\quad \tifeq{\tvar{i}}{\tsnd{\tvar{r}}+1}{\dint}{\\
-%	& \quad\quad\quad\quad\quad\quad \tifeq{\tvar{ts}}{\tnil{\kTypeBlur}}{\klist{\kTypeBlur}}{(\tden{\tvar{rx}}{\tvar{t}},\tvar{i})\\
-%	& \quad\quad\quad\quad\quad\quad}{(\tden{\titerm{\ifst{\iup{\tvar{rx}}}}}{\tvar{t}},\tvar{i})}\\	
-%	& \quad\quad\quad\quad\quad}{(\tden{\titerm{\isnd{\iup{\tvar{rx}}}}}{\tvar{t}},\tsnd{\tvar{r}}+1)}}{\terr}}}}}
-% }}})
-% }}{\terr}}}}}
-% }\\ & 
-%%& \quad \tops{pr}a}{b}{c}{d}
-% }{d}{i}{(\tlam{i}{\klist{\kTypeBlur}}{
-% 	\\& \quad \quad \tfold{\tvar{i}}{\titype{\dunit}}{s}{j}{r}{
-% 		\tfold{\tvar{j}}{\titype{\trepof{\tvar{s}}}}{\_}{\_}{\_}{\\
-%		&\quad\quad\quad
-% 		\titype{\dpair{\trepof{\tvar{s}}}{\dup{\tvar{r}}}}
-% 		}
-% 	}
-% })}
-%\end{flalign}
-%\caption{ABC}
-%\label{tuple}
-%\end{figure}
-%
-%\section{Limitations and Future Work}\label{limitations}
-%\todo{integrate this}Indeed, it is not a stretch to imagine a reasonably faithful implementation of our calculus starting with a pure, total subset of the term language of ML to construct the type-level language.
-We leave the development of such an implementation (and of bootstrapping type-level and internal languages that are themselves user-defined or extensible) as areas for future work to focus sharply on the theoretical foundations in this work.
-%
-%Because the input to an operator definition is a list of the recursively determined derivates of the argument in the same typing context as it appears in, our mechanism does not presently permit the definition of operator constructors that bind variables themselves or require a different form of typing judgement (e.g.
-additional forms of contexts).
-This is why the $\lambda$ form needs to be built in.
-It is the only form in the external language that can be used to bind variables.
-The type constructor $\fvar{Arrow}$ can, however, be defined from within the language and is included in the initial constructor context (it could be written out and included as a ``prelude'' as well).
-As seen in Fig.
-\ref{arrow}, it only defines the operator constructor $\opvar{ap}$.
-%
-%By writing the type assignment and translation logic in this way, as a type-level function, we are taking a rather ``implementation-focused view'' rather than attempting to extract a compiler from a declarative specification of a type system and an operational semantics.
-That is, @$\lambda$ leaves ultimately to the provider the problem of finding a decidable type assignment algorithm and resolve non-determinism that may arise.
-Similarly, we do not attempt to turn an operational semantics into a compiler automatically.
-We  also avoid difficulties with useful error reporting in practice.
-Simmons' discusses the difficulties of these issues \cite{rjs-princeton}.
-Composing two dialect specifications directly is more fraught with metatheoretic issues than analyzing two functional programs used together.
-By looking at the problem as a functional programmer, we can apply intuitions about abstraction barriers more directly.
-A solution that permits implementations to be extracted from fragment specifications could easily be layered atop this mechanism and inherit these guarantees.
-Alternatively, extension correctness could be mechanized by encoding these semantics in a proof assistant like Coq and then using its extraction mechanisms to produce a compiler.
-%
-%
-%The contexts and external term are ``input'' to the judgement, while the type assignment and translation are ``output'' (we briefly discuss potential variants in Sec.
-\ref{limitations}\todo{remove if not}).
-%%At best, they can aim to achieve a separation of concerns, such that each component of an application can be written using a specialized language (as advocated by proponents of the \emph{language-oriented approach} to software development \cite{language-oriented}), but there will still be an {interoperability} problem at the interface between components, as described above, unless an FFI has been explicitly developed for each pair of languages that might interface with each other, or library interfaces are constrained to only use a small subset of common constructs (as suggested in draft guidelines for F\#, for example \cite{fsharp-guidelines}) reducing  expressiveness and safety.
-%
-%  %We can specify the static semantics of the STLC as follows, using a uniform abstract syntax where type and operator indices are written in braces to emphasize this way of thinking about its structure:
-%%\begin{figure}[h]
-%%\small
-%%\vspace{-8pt}
-%%\begin{mathpar}
-%%\inferrule[var]{ }{
-%%	\jet{\eCtxX{\eCtx}{x}{\tau}}{x}{\tau}
-%%}
-%%
-%%\inferrule[arrow-I]{
-%%	\jet{\eCtxX{\eCtx}{x}{\tau}}{e}{\tau'}	
-%%}{
-%%	\jetX{\lambda[\tau]({x}.e)}{\fvar{Arrow}[(\tau, \tau')]}
-%%}
-%%
-%%\inferrule[arrow-E]{
-%%	\jetX{e_1}{\fvar{Arrow}[(\tau, \tau')]}\\
-%%	\jetX{e_2}{\tau}
-%%}{
-%%	\jetX{\texttt{ap}[()](e_1; e_2)}{\tau'}
-%%}
-%%\end{mathpar}
-%%\vspace{-8pt}
-%%\caption{Static semantics of the simply typed lambda calculus.}
-%%\end{figure}
-%
-%%\begin{figure}[h]
-%%\small
-%%\vspace{-8pt}
-%%\begin{mathpar}
-%%\inferrule[nat-I1]{ }{
-%%	\jetX{\texttt{z}[()]()}{\fvar{Nat}[()]}
-%%}
-%%
-%%\inferrule[nat-I2]{
-%%	\jetX{e}{\fvar{Nat}[()]}
-%%}{
-%%	\jetX{\texttt{s}[()](e)}{\fvar{Nat}[()]}
-%%}
-%%
-%%\inferrule[nat-E]{
-%%	\jetX{e_1}{\fvar{Nat}[()]}\\
-%%	\jetX{e_2}{\tau}\\
-%%	\jet{\eCtxX{\eCtxX{\eCtx}{x}{\fvar{Nat}[()]}}{y}{\tau}}{e_3}{\tau}
-%%}{
-%%	\jetX{\texttt{natrec}[()](e_1; e_2; x.y.e_3)}{\tau}
-%%}
-%%\end{mathpar}
-%%\vspace{-8pt}
-%%\caption{Static semantics of natural numbers.}\label{nat-statics}
-%%\vspace{-15pt}
-%%\end{figure}
-%%\begin{figure}[h]
-%%\small
-%%\begin{mathpar}
-%%\inferrule[Prod-I]{
-%%	\jetX{e_1}{\tau_1}\\
-%%	\jetX{e_2}{\tau_2}
-%%}{
-%%	\jetX{\texttt{pair}[()](e_1; e_2)}{\fvar{Prod}[(\tau_1, \tau_2)]}
-%%}
-%%
-%%\inferrule[Prod-E1]{
-%%	\jetX{e}{\fvar{Prod}[(\tau_1, \tau_2)]}
-%%}{
-%%	\jetX{\texttt{prl}[()](e)}{\tau_1}
-%%}
-%%
-%%\inferrule[Prod-E2]{
-%%	\jetX{e}{\fvar{Prod}[(\tau_1, \tau_2)]}
-%%}{
-%%	\jetX{\texttt{prr}[()](e)}{\tau_1}
-%%}
-%%\end{mathpar}
-%%\vspace{-8pt}
-%%\caption{Static semantics of products}\label{pair-statics}
-%%\end{figure}
-%
-%
-%%creating a new language.
-If this is not practical, the best one can attempt to do is encode the new types in terms of existing types (by a Church encoding, for example).
-This is generally unsatisfactory --
-%
-%%Languages implemented using these common patterns are central planning by a language designer or design committee. 
-%
-%%Researchers or domain experts who cannot work around such limitations must develop new standalone languages.
-In our simple scenario, we may simply copy our implementation of G\"odel's T or even edit it directly (a pernicious technique for implementing a new language where the prior one is overwritten).
-In a more complex scenario, we may instead employ a tool like a compiler generator or DSL framework \cite{fowler2010domain} that can generate a standalone implementation from declarative specifications of language constructs.
-Some of these tools allow you to package and reuse these specifications (with the important caveat that not all combinations of constructs are valid and free of conflicts, an important modularity issue that we will return to several times in this paper).
-%%
-%%The increasing sophistication and ease-of-use of these tools have led many to suggest a {\it language-oriented approach} \cite{journals/stp/Ward94} to software development where different components of an application are written in different languages.
-Unfortunately, this leads to problems at language boundaries: a library's external interface must only use constructs that can reasonably be expressed in \emph{all possible calling languages}.
-This can restrict domain-specific languages by, for example, precluding constructs that rely on statically-checked invariants stronger than those their underlying representation in a common target language normally supports.
-At best, constructs like these can be exposed by generating a wrapper where run-time checks have been inserted to guarantee necessary invariants.
-This compromises both verifiability and performance and requires the development of an interoperability layer for every DSL.
-Moreover, library clients must work with verbose and unnatural ``glue code'' when interfacing across languages, defeating the primary purpose of high-level programming languages: hiding the low-level details from the end-users of abstractions.
-We diagram this fundamental \emph{compatibility problem} in Figure \ref{approaches}(a).
-%%\begin{figure*}
-%%\begin{center}
-%%\includegraphics[scale=0.5]{approaches.pdf}
-%%\end{center}
-%%\vspace{-20px}
-%%\caption{\small (a) With a language-oriented approach, novel constructs are packaged into separate languages.
-Users can only safely and naturally call into languages consisting of common constructs (often only the common target language, such as C or Java bytecode).
-(b) With a language-internal extensibility approach, there is one system providing a common internal language, where additional primitive constructs that strictly strengthen its static guarantees or perform specialized code generation are specified and distributed within libraries.
-\label{approaches}}
-%%\end{figure*}
-%
-%%As a result, domain-specific languages and new general-purpose abstractions alike have experienced relatively slow adoption in practice.
-%%
-%%Porting large codebases to new languages is difficult, and the dominant programming languages innovate slowly, so programming language.
-%%
-%%More specifically, such languages are neither \emph{internally extensible} because the language itself exposes only natural numbers and functions to its users, nor are they \emph{externally extensible} because no new behaviors can be added to the language's  implementation in a separate module from the one containing the initial implementation.
-%
-%%This is the essence of a monolithic language implementation: it is impossible for anyone to modularly extend languages defined in this way. 
-%
-%%An extensible programming language could address these problems by providing a language-integrated mechanism for introducing new type and operator constructors and implementing their associated static and dynamic semantics directly.
-%%%Library developers need only consider which abstractions are most appropriate for their domain, without also considering whether these constructs can be exposed using abstractions appropriate to the domains of client code.
-Clients can simply import any necessary constructs when using a library that relies on them, preserving safety and ease-of-use without the use of  wrappers and glue code.
-We show this competing approach in Figure \ref{approaches}(b).
-%%%Researchers and domain experts thus gain the ability to distribute new ideas for evaluation to a broader development community without requiring the approval of maintainers of mainstream languages, large-scale porting of code or explicit interoperability layers.
-%%But, as mentioned in Section \ref{language-integrated-approaches}, some significant challenges must be addressed before such a mechanism can be relied upon.
-The desire for expressiveness must be balanced against  concerns about maintaining various safety properties in the presence of arbitrary combinations of user-defined  extensions to the language's core semantics.
-The mechanism must ensure that desirable \emph{metatheoretic properties} (e.g.
-type safety, decidability) of the language are maintained by extensions.
-Because multiple independently developed extensions might be used within one program, the mechanism must further guarantee their \emph{non-interference}.
-These are the issues we seek to address in this work.
-%
-%%When designing and implementing a new abstraction, experts typically begin by attempting to define new constructs in terms of existing language constructs.
-%%This approach is often effective because modern {general-purpose} abstraction mechanisms, like inductive datatypes and object systems, are highly expressive.  
-%%For example, the Delite framework leverages Scala's powerful general-purpose mechanisms to enable a number of useful \emph{embedded domain-specific languages} \cite{delite}.
-%%Unfortunately, there remain some situations of interest where general-purpose abstractions fall short. 
-%%For instance, it is difficult to adequately encode advanced type systems in terms of the simpler rules that govern general-purpose abstractions (e.g.
-reasoning about units of measure requires built-in language support in F\# \cite{conf/cefp/Kennedy09}).
-%%Even if a full encoding is possible, it may not be useful if it is overly verbose or unnatural, or if the error messages are overly abstract.
-For example, regular expressions encoded using inductive datatypes are quite verbose, so most functional languages support them via strings, which is less safe.
-%%Finally, general-purpose abstractions are implemented in a uniform manner, meaning domain-specific heuristics cannot be applied to eliminate overhead or perform optimizations, and implementations designed for typical application workloads may not be satisfactory in parts of a program where performance is a key criteria, such as when targeting heterogeneous hardware platforms (e.g.
-programmable GPUs) and distributed computing resources.
-%%
-%%% are at times impossible or impractical.
-In our example of adding products or sums to Godel's T, although Church encodings are possible \cite{pfpl}, they require a reasonable level of creativity\footnote{Anecdotally, Church encodings are among the more difficult-to-explain topics covered in our undergraduate programming languages course.}.
-Moreover, they will not offer the same static safety guarantees as a primitive encoding, they are more verbose and they will incur performance overhead by their use of closures rather than a more direct representation.
-This is not only a problem for simple languages like Godel's T.
-Several Haskell-based embedded DSLs have also needed to make significant compromises at times \cite{haskellDSLs}.
-{\color{red} examples?
-Scala?}
-%%
-%%%creating a new language.
-If this is not practical, the best one can attempt to do is encode the new types in terms of existing types (by a Church encoding, for example).
-This is generally unsatisfactory --
-%%
-%%%Languages implemented using these common patterns are central planning by a language designer or design committee. 
-%%
-%%Researchers or domain experts who run into situations like these, where more direct control over a language's semantics and implementation are needed, have little choice today but to realize new abstractions by creating a new language in some way.
-They might develop a new standalone language from scratch, modify an implementation of an existing language, or use tools like compiler generators, DSL frameworks and language workbenches \cite{fowler2010domain}.
-%%%In our simple scenario, we may simply fork our implementation of Godel's T or even edit it directly (a pernicious technique for implementing a new language where the prior one is overwritten). 
-%%%In a more complex scenario, we may instead employ a tool like a compiler generator or DSL framework \cite{fowler2010domain} that can generate a standalone implementation from declarative specifications of language constructs.
-Some of these tools allow you to package and reuse these specifications (with the important caveat that not all combinations of constructs are valid and free of conflicts, an important modularity issue that we will return to several times in this paper).
-%%The increasing sophistication and ease-of-use of these tools have led to calls for a {\it language-oriented approach} to software development, where different components of an application are written in different specialized languages \cite{journals/stp/Ward94}.
-Indeed, a number of software ecosystems are now designed explicitly to support many different languages, both general-purpose and domain-specific, atop a common intermediate language.
-The Java virtual machine (JVM), the Common Language Infrastructure (CLI) and LLVM are prominent examples of such ecosystems.
-%%
-%%Unfortunately, this leads to a critical problem at language boundaries: a library's external interface must only use constructs that can reasonably be expressed in \emph{all possible client languages}.
-This discourages languages from including constructs that rely on statically-checked invariants stronger than those supported by their underlying implementation in the common intermediate language.
-At best, constructs like these can be exposed by generating a wrapper where run-time checks have been inserted to guarantee these invariants.
-This compromises both verifiability and performance.
-%Forrequires the development of an interoperability layer for every pair of DSLs.
-%%Moreover, this approach exposes the internals of an implementation to clients, making the abstraction awkward to work with and causing code breakage when implementation details change.
-This defeats a primary purpose of high-level programming languages: hiding low-level details from clients of an abstraction.
-We diagram this fundamental \emph{interoperability problem} in Figure \ref{approaches}(a).
-%%%As an example, F\#'s type system prevents \lstinline{null} values from occurring within data structures, but because it's type system is not available when calling into F\# code from another language, like C\#, run-time null checks must still be included in the implementation.
-%%%\begin{figure*}[t]
-%%%\vspace{-15px}
-%%%\includegraphics[scale=0.415]{approaches.pdf}
-%%%\caption{(a) With the language-oriented approach, different primitive abstractions are packaged into separate languages that extend and target a common intermediate language (e.g.
-JVM bytecode).
-Users can only interface with libraries written in another language via the constructs in the common language, causing \emph{interoperability problems}.
-(b) With the language-internal approach, the semantics of new abstractions (i.e.
-the logic that governs typechecking and translation to a fixed {internal language}, here labeled \texttt{I}) can be implemented directly within so-called \emph{active libraries}.
-Clients can import and use these abstractions directly whenever needed.}
-%%%\label{approaches}
-%%%\end{figure*}
-%%
-%%%As a result, domain-specific languages and new general-purpose abstractions alike have experienced relatively slow adoption in practice.
-%%%
-%%%Porting large codebases to new languages is difficult, and the dominant programming languages innovate slowly, so programming language.
-%%%
-%%%More specifically, such languages are neither \emph{internally extensible} because the language itself exposes only natural numbers and functions to its users, nor are they \emph{externally extensible} because no new behaviors can be added to the language's  implementation in a separate module from the one containing the initial implementation.
-%%
-%%%This is the essence of a monolithic language implementation: it is impossible for anyone to modularly extend languages defined in this way. 
-%%
-%%%Programming languages are typically designed around a monolithic collection of primitive type families and operators.
-Consider, as a simple example, Godel's T \cite{pfpl}, a typed lambda calculus with recursion on primitive natural numbers\todo{add statics to Appendix A}.
-Although a language designer may casually speak of ``extending Godel's T with primitive product and sum types'', adding these type families and associated operators to this kind of language from within is impossible.
-That is, Godel's T is not \emph{internally extensible}.
-%%
-%%\emph{Internally-extensible programming languages} promise to avoid these problems by providing researchers and domain experts with a mechanism for implementing the semantics of new primitive constructs directly within libraries.
-%%%Developers of libraries need only determine whether they are appropriate for their domain, without also considering whether these constructs can be exposed in terms of abstractions appropriate to client code.
-%%As a result, clients can granularly import any necessary primitive constructs when using code that relies on them, and thereby achieve full safety, ease-of-use and performance.
-Providers of components thus need only consider whether primitives that they use are appropriate for their domain, without also considering whether their code might be used in a context where these primitives are not otherwise appropriate.
-Libraries containing logic that is invoked at compile-time, as extension logic would be, have been called \emph{active libraries} \cite{activelibraries}.
-We adopt this terminology and diagram this competing approach in Figure \ref{approaches}(b).
-%%
-%%%Researchers and domain experts thus gain the ability to distribute new ideas for evaluation to a broader development community without requiring the approval of maintainers of mainstream languages, large-scale porting of code or explicit interoperability layers.
-%%
-%%For a language-internal extension mechanism to be feasible, however, it must achieve expressiveness while also ensuring that extensions cannot compromise the safety properties of the language and its tools, nor interfere with one another.
-That is, extensions cannot simply be permitted to add arbitrary logic to the type system or compiler, because this would make it possible to break  type safety, decidability or adequacy theorems that are critical to the operation of the language, the compiler or other extensions.
-We review some previous attempts at language extensibility, and highlight how they do not adequately achieve both safety and expressiveness, in Section \ref{related-work}.
-%%%{\color{red} transition here} Correctness properties of an extension itself should be modularly verifiable, so that its users can rely on it for verifying and compiling their own code.
-The mechanism must also ensure that desirable metatheoretic properties and global safety guarantees of the language cannot be weakened by extensions.
-And with multiple independently-developed extensions used at once, the mechanism must further guarantee that they cannot interfere with one another.
-%%
-%%In this paper, we introduce a language-internal extensibility mechanism called \emph{active typechecking and translation} (AT\&T) that allows developers to introduce and implement the logic governing new primitive type and operator families from within libraries.
-%%We argue that this can be accomplished by enriching the type-level language, rather than introducing a separate metalanguage into the system. 
-%%To make this proposal concrete, we begin by introducing a simple core calculus, called \atlam~(for the ``actively-typed lambda calculus''), in Section \ref{atlam}. 
-%%This calculus uses type-level computation of higher kind, along with techniques borrowed from the typed compilation literature and a form of type abstraction that ensures that the implementation details of an extension are not externally visible to guarantee the safety of the language, the decidability of typechecking and compilation and non-interference of extensions, as we outline in Section \ref{safety}.
-%%In Section \ref{examples}, we suggest that despite these constraints, this mechanism is expressive enough to admit, within libraries, a number of general-purpose and domain-specific abstractions that normally require built-in language support.
-%%
-%%Our core calculus uses a uniform abstract syntax for primitive operators to simplify our presentation and analysis, but this syntax is too verbose to be practical.
-Thus, we begin this section by showing how a key design choice made in the calculus -- to associate operators with type families, forming what we call \emph{active type families} -- supports a novel type-directed desugaring mechanism that permits the use of conventional concrete syntax for language extensions.
-%
-%% Our choice of a simply-typed, simply-kinded calculus where expressions are given meaning by translation to a simply-typed internal language appears to occupy a ``sweet spot'' in the design space, and relates closely to how simply-typed functional languages like  ML and Haskell are specified and implemented today.
-In Section \ref{design}, we briefly discuss other points in the design space of actively-typed languages and describe the sorts of abstractions that the mechanism as we have introduced it is not capable of expressing, suggesting several directions for future research.
-We conclude with a discussion of related work in Section \ref{related-work}.
-%%dependently-typed and object-oriented type-level languages, as well as the constraints governing the design of the internal language.  
-%%We will also note how object-oriented techniques may also be suit, because of a fundamental connection to the \emph{expression problem} \cite{expression-problem}.
-%
-%%specify new typechecking rules and translation logic from within libraries.
-The AT\&T mechanism utilizes type-level computation of higher kind and integrates typed compilation techniques into the language to provide strong safety guarantees, while remaining straightforward and expressive.
-%
-%%AT\&T is general with respect to many choices about the type-level language, the typed internal language and syntax.
-Choices along these dimensions can affect both expressiveness and ease-of-use.
-We will begin in Sec.
-2 by introducing a minimal system called $@\lambda$ (the ``actively-typed lambda calculus'') that distills the essence of the mechanism in a simply-typed, simply-kinded setting.
-This will allow us to fully and concisely formalize the language and compiler and give several key safety theorems.
-We will then continue in Sec.
-3 by discussing variants of this mechanism based on other basic paradigms, considering dependently-typed functional languages and object-oriented languages, discussing trade-offs between expressivity and safety when doing so.
-We have developed a simple prototype called Ace and have used it to develop a number of full-scale language extensions as libraries.
-We will briefly discuss this language and these extensions in Sec.
-4.
-%
-%%We note at the outset that AT\&T focuses on extending the static semantics of languages with fixed, though flexible, syntax.
-Language-internal syntax extension mechanisms have been developed in the past (e.g.
-SugarJ \cite{sugarj}) but they have also suffered from safety problems because grammar composition is not always safe when done in an  unconstrained manner.
-Constrained approaches that provide stronger safety guarantees have recently been outlined (e.g.
-Wyvern \cite{globaldsl13}) but we will leave integration of syntax extensions with semantic extensions as future work.
-%%\section{From Extensible Compilers to Extensible Languages}\label{evolution}
-%%\begin{figure}[t]
-%%\small
-%%$$\begin{array}{rccl}	
-%%\textbf{programs} & \rho & ::= & \pfam{\familyDf}{\progsort}  \\
-%% & & \pipe &  \pdef{t}{\kappa}{\tau}{\progsort} \pipe e\\
-%%\text{primitive ops}		&	\theta	&	::= &	\tops{op}{\kappaidx}{i}{a}{\tau} \pipe 
-%%												\topp{\theta}{\theta}\\
-%%\\
-%%\textbf{external terms} 				&	e	&	::=	&	\evar{x} \pipe 
-%%														\elam{\evar{x}}{\tau}{e} \pipe 
-%%														\eop{Fam}{op}{
-%%															\taui
-%%														}{
-%%  												    		\splat{e}{1}{n}
-%%														} \\
-%%									& 		&		& 	\\
-%%							
-%%\hspace{-5pt}\textbf{type-level terms} 	& \tau 	& ::= 	& 	\tvar{t} \pipe 
-%%														\tlam{t}{\kappa}{\tau} \pipe 
-%%														\tapp{\tau_1}{\tau_2}\pipe
-%%														\iintlit \pipe \iop{\tau_1}{\tau_2} \pipe \tstr{str} \\
-%%									    & &  \pipe & 
-%%														\tnil{\kappa} \pipe \tcons{\tau_1}{\tau_2} \pipe 
-%%									                     \tfold{\tau_1}{\tau_2}{h}{t}{r}{\tau_3}
-%%														\\
-%%												
-%%	 			& 		& \pipe	& 	 \tunit \pipe 
-%%														\tpair{\tau_{1}}{\tau_{2}} \pipe 
-%%														\tfst{\tau} \pipe 
-%%														\tsnd{\tau} 
-%%														\\
-%%\text{equality}  & & \pipe & 					\tifeq{\tau_{1}}{\tau_{2}}{\kappa}{\tau_{3}}{\tau_{4}} 
-%%														\\														
-%%
-%%\text{types} 						& 		& \pipe	& 	\ttypestd \\
-%% & & \pipe &  \tfamcase{\tau}{Tycon}{x}{\tau_1}{\tau_2}\\
-%%						%				& & \pipe & \tfamcase{\tau}{Fam}{x}{\tau_1}{\tau_2}\\
-%%																								
-%%\text{denotations} 				& 		 & 	\pipe	&	\tden{\tauiterm}{\tautype} \pipe \terr \\
-%% & & \pipe &  \tdencase{\tau}{x}{t}{\tau_1}{\tau_2}\\
-%% %& & \pipe & 
-%%%														\tdencase{\tau}{y}{x}{\tau_1}{\tau_2}
-%%%														 \\
-%%
-%%\text{reified IL}		&		&	\pipe	&	\titerm{\iota} \pipe \titype{\sigma} \\
-%%
-%%												\\
-%%\textbf{kinds} 					& \kappa	&	::=	&	\karrow{\kappa_1}{\kappa_2} \pipe \klist{\kappa} \pipe \dint \pipe
-%%											    \kstr \pipe
-%%												\kunit \pipe 
-%%												\kpair{\kappa_{1}}{\kappa_{2}} \\
-%%	& & \pipe &  
-%%												\kTypeBlur \pipe \kDen \pipe 
-%%												\kIType \pipe \kITerm
-%%												\\
-%%\\												
-%%%\textbf{ops signature}			& \Theta	&	::=	&	\kOpEmpty \pipe \kOp{\Theta}{op}{\kappai}\\
-%%%											 							&		&		&	\\
-%%\textbf{internal terms} 				& 	\iota	&	::=	&	\evar{x} \pipe 
-%%												\ilam{\evar{x}}{\sigma}{\iota} \pipe 
-%%												\iapp{\iota_{1}}{\iota_{2}} \pipe
-%%												\ifix{\evar{f}}{\sigma}{\iota} \\
-%%									& & \pipe & 
-%%												\ipair{\iota_{1}}{\iota_{2}} \pipe 
-%%												\ifst{\iota} \pipe
-%%												\isnd{\iota}  
-%%												\\
-%%							& 		& 	\pipe	& 
-%%												\iintlit \pipe \iop{\iota_{1}}{\iota_{2}} \pipe \iIfEq{\iota_{1}}{\iota_{2}}{\dint}{\iota_{3}}{\iota_{4}}  
-%%												\\
-%% &  & \pipe & \tvalof{\tau_1}{\tau_2} \pipe \iup{\tau} \\
-%%%\text{deabstracted}& \iota & ::= & \mathcal{G}[\iota, \sigma]\\
-%%\textbf{internal types}			&	\sigma	&	::=	&    \darrow{\sigma_1}{\sigma_2} \pipe
-%%												\dint \pipe
-%%												\dpair{\sigma_1}{\sigma_2} \pipe
-%%												\trepof{\tau} \pipe \dup{\tau}\\
-%%\end{array}$$
-%%%\vspace{-10pt}
-%%\caption{\small Syntax of \atlam.
-Variables $x$ are used in expressions and internal terms and are distinct from type-level variables, $\tvar{t}$.
-Names $\fvar{Fam}$ are type family names (we assume that globally unique type family names can be generated by some external mechanism) and $\opvar{op}$ are operator family names.
-$\tstr{str}$ denotes string literals, $\iintlit$ denotes integer literals and $\oplus$ stands for binary operations over integers.
-%The productions related to the internal language are written using generators $\mathcal{G}$ and $\mathcal{S}$ to avoid duplicating the syntax of common terms.
-%%\label{grammar}}
-%%\end{figure}
-%%\begin{figure}[t]
-%%\small
-%%\begin{flalign}
-%%\label{natfam}&\family{Nat}{\kunit}{\\
-%%\label{z}&\quad\tops{z}{\kunit}{i}{a}{
-%%	\tapp{\tapp{\tvar{empty}}{\tvar{a}}}{
-%%		\tden{\titerm{0}}{\ttype{Nat}{\tunit}}
-%%	}};\\
-%%\label{s}&\quad\tops{s}{\kunit}{i}{a}{
-%%	\tapp{
-%%		\tapp{\tvar{pop\_final}}{\tvar{a}}}{\tlam{x}{\kITerm}{\tlam{t}{\kTypeBlur}{
-%%			\\
-%%			&\quad\quad\tapp{\tapp{\tapp{\tvar{check\_type}}{\tvar{t}}}{\ttype{Nat}{\tunit}}}{
-%%				\tden{\titerm{\iup{\tvar{x}}+1}}{\ttype{Nat}{\tunit}}
-%%			}
-%%		}}
-%%		}
-%%	};\\
-%%\label{rec}&\quad\tops{rec}{\kunit}{i}{a}{
-%%	\tapp{\tapp{\tvar{pop}}{\tvar{a}}}{\tlam{x1}{\kITerm}{\tlam{t1}{\kTypeBlur}{\tlam{a}{\klist{\kDen}}{\\
-%%	&\quad\quad\tapp{\tapp{\tvar{pop}}{\tvar{a}}}{\tlam{x2}{\kITerm}{\tlam{t2}{\kTypeBlur}{\tlam{a}{\klist{\kDen}}{\\
-%%	&\quad\quad \tapp{\tapp{\tvar{pop\_final}}{\tvar{a}}}{\tlam{x3}{\kITerm}{\tlam{t3}{\kTypeBlur}{\\
-%%	&\quad\quad \tapp{\tapp{\tapp{\tvar{check\_type}}{\tvar{t1}}}{\ttype{Nat}{\tunit}}}{(\\
-%%	&\quad\quad \tapp{\tapp{\tapp{\tvar{check\_type}}{\tvar{t3}}}{\ttype{Arrow}{(\ttype{Nat}{\tunit},\ttype{Arrow}{(\tvar{t2},\tvar{t2})})}}}{\\
-%%		\label{fix}&\quad\quad \tden{\titerm{\iapp{(\ifix{f}{\darrow{\dint}{\trepof{\tvar{t2}}}}{\ilam{x}{\dint}{\\
-%%		\label{lastop}&\quad\quad\quad \iIfEq{x}{0}{\dint}{\iup{\tvar{x2}}}{\iapp{\iapp{\iup{\tvar{x3}}}{(x-1)}}{(\iapp{f}{(x-1)})}}}})}{\iup{\tvar{x1}}}}}{\tvar{t2}}
-%%	)}}
-%%	}}}}}}}}}}}}
-%%\\
-%%&}{XXX}{i}{\titype{\dint}};\\
-%%\label{nattype}&\pdef{nat}{\kTypeBlur}{\ttype{Nat}{\tunit}}{\\
-%%\label{plus}&(\elam{plus}{\ttype{Arrow}{(\tvar{nat}, \ttype{Arrow}{(\tvar{nat}, \tvar{nat}))}}}{
-%%	\elam{two}{\tvar{nat}}{\\
-%%\label{ap}	&\quad\quad \eopapp{\eopapp{plus}{two}}{two}}})
-%%}\\
-%%\label{add}&\quad (\elam{x}{\tvar{nat}}{\elam{y}{\tvar{nat}}{
-%%	\eop{Nat}{rec}{\tunit}{x; y; \elam{p}{\tvar{nat}}{\elam{r}{\tvar{nat}}{
-%%	\eop{Nat}{s}{\tunit}{r}
-%%	}}}}})\\
-%%\label{two}&\quad  \eop{Nat}{s}{\tunit}{\eop{Nat}{s}{\tunit}{\eop{Nat}{z}{\tunit}{ }}}
-%%%{\\&\elet{two}{\tvar{nat}}{\eop{Nat}{s}{\tunit}{\eop{Nat}{s}{\tunit}{\eop{Nat}{z}{\tunit}{ }}}}{\eapp{\eapp{plus}{two}}{two}}}
-%%%}
-%%%\elam{x}{\tvar{nat}}{\elam{y}{\tvar{nat}}{\\
-%%	%&\quad \eop{Nat}{rec}{\tunit}{x; y; \elam{p}{\tvar{nat}}{\elam{r}{\tvar{nat}}{
-%%	%\eop{Nat}{s}{\tunit}{r}
-%%	%}}}}
-%%\end{flalign}
-%%\caption{G\"odel's T in \atlam, used to calculate 2+2.
-Helper functions for working with lists ($\tvar{empty}$, $\tvar{pop}$, $\tvar{pop\_final}$) and types ($\tvar{check\_type}$),  described below, are given in the appendix.
-We will refer to this program as $\rho_{\text{nat}}$ in the text.}
-%%\label{example}
-%%\end{figure}
-%%
-%%To understand the genesis of our internal extension mechanism, it is helpful to begin by considering why most implementations of programming languages cannot even be  externally extended. 
-%%Let us consider, as a simple example, an implementation of G\"odel's T, a typed lambda calculus with recursion on primitive natural numbers (see Appendix). 
-%%A compiler for this language written using a functional language will invariably represent the primitive type families and operators using {closed} inductive datatypes. 
-%%For example, a simple implementation in Standard ML may be based around these datatypes:
-%%\begin{lstlisting}
-%%  datatype Type = Nat | Arrow of Type * Type
-%%  datatype Exp = Var of var 
-%%               | Lam of var * Type * Exp | Ap of Exp * Exp 
-%%               | Z | S of Exp | Natrec of Exp * Exp * Exp
-%%\end{lstlisting}
-%%
-%%The logic governing typechecking and translation to a suitable intermediate language (for subsequent optimization and compilation by some back-end) will proceed by exhaustive case analysis over the constructors of \lstinline{Exp}.
-%%
-%%In an object-oriented implementation of Godel's T, we might instead encode types and operators as subclasses of abstract classes \lstinline{Type} and \lstinline{Exp}.
-Typechecking and translation will proceed by the ubiquitous \emph{visitor pattern}  by dispatching against a fixed collection of {known} subclasses of \lstinline{Exp}.
-%%
-%%In either case, we encounter the same basic issue: there is no way to modularly add new primitive type families and operators and implement their associated typechecking and translation logic. 
-%%%This issue is related to the widely-discussed \emph{expression problem} (in a restricted sense -- we do not consider adding new functions beyond typechecking and translation here, only adding logic to these) \cite{wadler-expression}.
-%%
-%%A number of language mechanisms have been proposed that allow new cases to be added to datatypes and the functions that operate over them in a modular manner. 
-%%In functional languages, we might use \emph{open datatypes}.
-For example, if we wish to extend G\"odel's T with product types and we have written our compiler in a language supporting open inductive datatypes, it might be possible to add new cases like this:
-%%\begin{lstlisting}
-%%  newcase Prod of Type * Type extends Type
-%%  newcase Pair of Exp * Exp extends Exp    (* Intro *)
-%%  newcase PrL of Exp extends Exp           (* Elim Left *)
-%%  newcase PrR of Exp extends Exp           (* Elim Right *)
-%%\end{lstlisting}
-%%
-%%The logic for functionality like typechecking and translation could then be implemented for only these new cases.
-For example, the \lstinline{typeof} function that assigns a type to an expression could be extended like so:
-%%\begin{lstlisting}
-%%  typeof PrL(e) = case typeof e of 
-%%      Prod(t1, _) => t1 
-%%    | _ => raise TypeError("<appropriate error message>")
-%%\end{lstlisting}
-%%
-%%If we allowed users to define new modules containing definitions like these and link them into our compiler, we will have succeeded in creating an externally-extensible compiler, albeit one where safety is not guaranteed (we will return to this point shortly).
-We have not, however, created an extensible programming language, for two reasons.
-First, compiler extensions are distributed and activated separately from libraries, so dependencies become more difficult to manage.
-Second, other compilers for the same language will not necessarily support the same extensions.
-%%If our newly-introduced constructs are exposed at a library's  interface boundary, clients using different compilers face the same problems with interoperability that those using different languages face.
-That is, {extending a language by extending a single compiler for it is morally equivalent to creating a new language}.
-Several prominent language ecosystems today are in a state where a prominent compiler has introduced or enabled the introduction of extensions that many libraries have come to rely on, including the Glasgow Haskell Compiler, SML/NJ and the GNU compilers for C and C++.
-%%
-%%A more appropriate and useful place for extensions like this is directly within libraries, alongside abstractions that can be adequately implemented in terms of existing primitive abstractions.
-To enable this, the language must allow for the introduction new primitive type families, like \lstinline{Prod}, operators, like \lstinline{Pair}, \lstinline{PrL} and \lstinline{PrR}, and associated typechecking and translation logic.
-When encountering these new operators in expressions, the compiler must effectively  hand control over typechecking and translation to the appropriate user-defined logic.
-Because this mechanism is {language-internal}, all compilers must support it to satisfy the language specification.
-%%
-%%Statically-typed languages typically make a distinction between \emph{expressions}, which describe run-time computations, and type-level constructs like types, type aliases and datatype declarations.
-The design described above suggests we may now need to add another layer to our language, an {extension language}, where extensions can be declared and implemented.
-In fact, we will show that \textbf{the most natural place for type system extensions is within the type-level language}.
-The intuition is that extensions to a statically-typed language's semantics will need to manipulate types as values at compile-time.
-Many languages already allow users to write type-level functions for various reasons, effectively supporting this notion of types as values at compile-time (see Sec.
-\ref{related-work} for examples).
-The type-level language is often constrained by its own type system (where the types of type-level values are called \emph{kinds} for clarity) that prevents type-level functions from causing problems during compilation.
-This is precisely the structure that a distinct extension layer would have, and so it is quite natural to unify the two, as we will show in this work.
-%%
-%%\section{\atlam}\label{atlam}
-%%In this section, we will develop a core calculus, called @$\lambda$ for the ``actively-typed lambda calculus'', by way of a semantics and a simple example, and discuss how it addresses the safety concerns that arise.
-%%\subsection{Overview}
-%%The grammar of \atlam~is shown in Figure \ref{grammar}. 
-%%A program, $\rho$, consists of a series of declarations followed by an expression.
-Declarations can be either bindings of type-level terms to type-level variables using \textsf{def} or a primitive type family declared using \textsf{family}, which can contain implementations of one or more operator families, $\theta$.
-Expressions, $e$, can be either variables, lambdas, or applications of operators.
-%%%, and are ultimately given meaning by translation to a typed internal language, with terms $\iota$ and types $\sigma$. 
-%%%This language has been chosen, for simplicity, to be a variant of Plotkin's PCF with primitive integers and products, but in practice would include other constructs consistent with its role as a high-level intermediate language.
-%%
-%%The language is structured as a simply-typed lambda calculus with simply-kinded type-level computation. Kinds, $\kappa$, classify type-level terms, $\tau$. 
-%%Types are type-level values of kind $\star$ (following System $F_{\omega}$) and classify expressions.
-The type-level language also includes other kinds of terms: type-level functions, lists (required by our mechanism), integers, strings and products for the sake of our examples (see Sec.
-\ref{safety} for a discussion on other acceptable kinds of type-level data) and constructs for developing extensions -- denotations and reified internal terms and types -- which we will discuss in the sections below.
-%%
-%%All expressions are given meaning by translation to a typed internal language.
-This language has been chosen, for simplicity, to be a variant of Plotkin's PCF with primitive integers and products, but in practice would include other constructs consistent with its role as a high-level intermediate language.
-The grammar of internal terms, $\iota$, and internal types, $\sigma$, also includes special forms containing type-level terms; these are used for developing extensions and during compilation and will be erased before compilation ends.
-%%\subsection{Example: G\"odel's T as an Active Type Family}
-%%To make our explanation of each of the constructs in the calculus concrete, we will work through an example showing how to introduce primitive natural numbers with bounded recursion in the style of G\"odel's T \cite{pfpl}.
-These will be implemented internally as integers (that is, internal terms of internal type $\dint$).
-Figure \ref{example} shows how to define the indexed type family $\fvar{Nat}$.
-This family contains only one type, written $\ttype{Nat}{\tunit}$, which we alias on line \ref{nattype} by defining the type-level variable $\tvar{nat}$.
-We define the typechecking and translation logic for the operators associated with this family (\opvar{z}, \opvar{s} and \opvar{rec}) on lines \ref{z}-\ref{lastop} and use these to define a $plus$ function on line \ref{add} and compute 2+2.
-We will refer back to this figure as we describe each construct below.
-%%
-%%\subsection{Indexed Type Families and Types}\label{families}
-%%The syntactic form $\familyDf$ declares a new primitive type family named $\fvar{Fam}$ indexed by type-level values of kind $\kappaidx$ with representation schema $\tvar{i}.\taurep$ and operators $\theta$.
-The purpose of the representation schema and of associating of operators directly with types will be explained below.
-%%
-%%Declaring a type family in this way is a language-internal analog to adding a new constructor to the compiler-internal datatype \lstinline{Type}, as suggested in Sec. \ref{evolution}. 
-%%The index represents the data associated with this constructor.
-A type (that is, a type-level term of kind $\kTypeBlur$) is constructed by naming a family in scope and providing a type-level term of the appropriate kind as an index.
-A base type like $\tvar{nat}$ can be thought of as being the only type in the family $\fvar{Nat}$ trivially indexed by the unit value, of kind $\kunit$, while families like $\fvar{Ntuple}$ might be indexed by a list of types, having kind $\klist{\kTypeBlur}$.
-%% For example, $\ttype{Ntuple}{\tcons{\tvar{nat}}{\tcons{\tvar{nat}}{\tnil{\kTypeBlur}}}}$ might be the type of a pair of natural numbers.
-Given a type, its family can be \textsf{case} analyzed to extract the value of its index.
-It is important that type equality be decidable, so only kinds for which equivalence coincides with syntactic equality can be used as type family indices.
-The main  consequence of this restriction is that indices cannot contain type-level functions.
-%%
-%%\subsection{Representations and Representation Schemas}
-%%As we will discuss further below, it is important that all expressions classified by a type compile to consistently-typed internal terms.
-For this reason, we require that every type have a single internal type associated with it, called its \emph{representation}.
-This is computed by substituting the type index for the bound variable $\tvar{i}$ in the term $\taurep$, called the \emph{representation schema} of the type family, and evaluating to a value representing an internal type.
-Internal types, $\sigma$, are reified as type-level terms of kind $\kIType$ using the introductory form $\titype{\sigma}$.
-%%
-%%\subsection{Indexed Operator Families and Denotations}\label{operators}
-%%Type families are also equiped with a collection of primitive operator families.
-An operator family named $\opvar{op}$ is declared using the form $\tops{op}{\kappaidx}{i}{a}{\tauop}$.
-Like type families, operator families are indexed by values of some kind, $\kappaidx$, but because operators are not first-class type-level values in our calculus, there are no equality restrictions.
-In the example in Fig.
-\ref{example}, all the operators are trivially indexed by the kind $\kunit$, so each family only contains one operator.
-However, a type family like $\fvar{Ntuple}$ would be equipped with a family of projection operators, $\opvar{pr}$, indexed by a position (e.g.
-an integer in our calculus).
-A family implementing record types or object types might have a similar operator indexed by a type-level string representing the field being accessed (see Sec.
-\ref{examples}).
-%%
-%%To apply an operator, the grammar provides a uniform form of expression: $\eop{Fam}{op}{\tauidx}{e_1; \ldots; e_n}$, where $n \geq 0$.
-The typechecking and translation of an expression of this form is controlled by the term $\tauop$ in the operator's declaration.
-This term must evaluate to a \emph{denotation}, which is a type-level value of kind $\kDen$, when given the operator index, $\tauidx$, and a list constructed from the denotations recursively assigned to each argument, $e_1$ through $e_n$.
-%%
-%%There are two forms of denotations that an expression can be assigned.
-A \emph{valid denotation} has the form $\tden{\tauiterm}{\tautype}$, where $\tauiterm$ is the \emph{translation} of the expression to an internal term and $\tautype$ is the type it has been assigned.
-Internal terms are represented as type-level terms using the form $\titerm{\iota}$, and have kind $\kITerm$ (similar to $\kIType$).
-%It is important to note that there are no elimination form for reified terms or types.
-%%%(indeed, in \atlam, terms are never  syntactically analyzed directly by extensions, unlike macro systems and other forms of term rewriting; see Sec. \ref{related-work})
-%%If a type error is detected by an operator, the \emph{error denotation}, $\terr$, is returned instead of a valid denotation.
-In a practical implementation, a specialized error message and other diagnostic information would be provided when returning $\terr$, but we omit such details for simplicity.
-Terms of kind $\kDen$ can be \textsf{case} analyzed to determine if they are valid or errors, and if valid, to extract the translation and type.
-%%
-%%In the example in Fig.
-\ref{example}, the operator $\opvar{z}$ checks that no arguments were passed in using the simple helper function $\tvar{empty}$.
-If so, it returns a valid denotation by pairing the translation $\titerm{0}$ with the type $\ttype{Nat}{\tunit}$, as expected\footnote{Actually, there is no theoretical barrier to a different ``zero'' being used!}.
-If an argument was provided, the helper function returns $\terr$.
-The successor operator, $\opvar{s}$, takes one argument, so it pops a denotation off the argument list, making sure there are no more, and binds its translation and type to $\tvar{x}$ and $\tvar{t}$ respectively, all using the helper function $\tvar{pop\_final}$.
-It then checks that the argument's type is also $\ttype{Nat}{\tunit}$, returning a denotation pairing the translation $\titerm{\iup{\tvar{x}} + 1}$ with the type $\ttype{Nat}{\tunit}$ if so.
-The form $\iup{\tau}$ is used to ``un-reify'' reified internal terms, of kind $\kITerm$ (thus serving as the left-inverse of $\titerm{\iota}$).
-In this case, $\tvar{x}$ is a type-level variable of kind $\kITerm$ representing the translation of the argument to the successor operator, so we simply need to add one to it.
-Because we have checked that the denotation's type was $\ttype{Nat}{\tunit}$, and the representation schema will guarantee that expressions of this type always translate to integers, we know that it is safe to do so.
-If any of these steps fail, the various helper functions we use simply return $\terr$ (in practice, it would be prudent to equip each failure condition with a different error message).
-Compilation will also fail if we accidentally violate the representation schema, as we will see in Sec.
-\ref{repcon}.
-%%
-%%% The operator index, $\tauidx$, and . 
-%%
-%%\subsection{Functions, Variables and Arrow Types}
-%%
-%%The recursor operator, $\opvar{rec}$, proceeds similarly, extracting the translations and types of each of its three arguments.
-Of note, however, is how it handles the third argument, which binds two variables (the predecessor and the result of recursing on it).
-In \atlam, the built-in $\lambda$ operator serves as the sole mechanism for introducing bound variables.
-In most calculi, lambda terms have types of the form $\tau_1 \rightarrow \tau_2$.
-In \atlam, $\rightarrow$ corresponds to the built-in type family, $\fvar{Arrow}$.
-This family is indexed by a pair of types, $\kpair{\kTypeBlur}{\kTypeBlur}$, and is always in scope.
-Its representation schema simply maps to the corresponding internal arrow type, $\tvar{i}.\titype{\darrow{\trepof{\tfst{\tvar{i}}}}{\trepof{\tsnd{\tvar{i}}}}}$.
-The special form $\trepof{\tau}$ is used to refer, abstractly, to the representation of the type $\tau$ (we also see this used on line \ref{fix}).
-Although the $\lambda$ operator is built-in, because it needs to bind variables, application is just an operator associated with the $\fvar{Arrow}$ family, $\opvar{ap}$, as is seen on line \ref{ap}.
-The details are straightforward and given in the appendix.
-%%
-%%\subsection{Compilation}
-%%\begin{figure}[t]
-%%\small
-%%\begin{mathpar}
-%%\inferrule{
-%%	\overbrace{\progOK{\emptyctx}{\fvalCtx_0}{\rho}}^{\text{\normalsize Kind Checking}}\\
-%%	\overbrace{\pcompiles{\fvalCtx_0}{\rho}{\iota}}^{\text{\normalsize Active Typechecking and Translation}}
-%%}{
-%%	\ptcc{\rho}{\iota}
-%%}
-%%\end{mathpar}
-%%\caption{\small Central Compilation Judgement of \atlam.}
-%%\label{ccj}
-%%\end{figure}
-%%The \emph{central compilation judgement}, shown in Figure \ref{ccj}, captures the two phases of compilation: kind checking and active typechecking and translation. 
-%%The first phase ensures that all type-level terms in the program are well-kinded and that all expressions and internal terms are closed.
-The kinding rules for programs are given in Figure \ref{kindprog}, and they rely on the kinding rules for type-level terms given in Figure \ref{tlkind}.
-These rules use contexts $\Sigma$ and $\Theta$ to track family and operator signatures, but beyond that, the rules are largely consistent with those of a simply-typed lambda calculus shifted into the type-level, with the addition of the handful of special forms constrained as described in the previous sections.
-The reader is encouraged to verify that the example in Fig.
-\ref{example} is well-kinded.
-%%% As we will see, the kind checking phase ensures that many kinds of errors are ruled out and that this process will not ``get stuck'' (see Sec.
-\ref{safety}).
-The evaluation semantics for type-level terms are given in Fig.
-\ref{tleval}.
-%%\begin{figure}[t]
-%%\small
-%%$\fbox{\inferrule{}{\progOKX{\progsort}}}$
-%%~~~$\tvarCtx ::= \emptyctx \pipe \tvarCtxX{t}{\kappa}$~~~
-%%$\fCtx ::= \Sigma_0	 \pipe \fvalCtxX$
-%%\begin{mathpar}
-%%\inferrule[family-kinding]{
-%%	\fvar{Fam} \notin \text{dom}(\fvalCtx)\\
-%%	\kEq{\kappaidx}\\
-%%	\opType{\tvarCtx}{\fvalCtxX}{\theta}{\Theta}\\\\
-%%	\tKind{\tvarCtx, \tvar{i}:{\kappaidx}}{\fvalCtxX}{\tau}{\kIType}\\
-%%	\progOK{\tvarCtx}{\fvalCtxX}{\rho}
-%%}{
-%%	\progOKX{\pfam{\familyDf}{\rho}}
-%%}
-%%
-%%\inferrule[def-kinding]{
-%%	\tKindX{\tau}{\kappa}\\
-%%	\progOK{\tvarCtxX{t}{\kappa}}{\fvalCtx}{\rho}
-%%}{
-%%	\progOKX{\pdef{t}{\kappa}{\tau}{\rho}}
-%%}
-%%
-%%\inferrule[exp-kinding]{
-%%	\exprOK{\tvarCtx}{\emptyctx}{\fvalCtx}{e}
-%%}{
-%%	\progOKX{e}
-%%}
-%%\end{mathpar}
-%%$\fbox{$\tKindX{\theta}{\Theta}$}$
-%%~~~$\Theta ::= \kOpS{op}{\kappaidx} \pipe \Theta, \Theta$~~~
-%%\begin{mathpar}
-%%\inferrule[op-kinding]{
-%%	\tKind{\tvarCtx, \tOfKind{\tvar{i}}{\kappai}, \tOfKind{\tvar{a}}{\klist{\kDen}}}{\fvalCtx}{\tau}{\kDen}
-%%}{
-%%	\opType{\tvarCtx}{\fvalCtx}{{\tops{op}{\kappaidx}{i}{a}{\tau}}}{
-%%	\kOpS{op}{\kappaidx}}
-%%}
-%%
-%%\inferrule[ops-kinding]{
-%%	\tKindX{\theta_1}{\Theta_1}\\
-%%	\tKindX{\theta_2}{\Theta_2}\\\\
-%%	\text{dom}(\theta_1) \cap \text{dom}(\theta_2) = \emptyset
-%%}{
-%%	\tKindX{\theta_1; \theta_2}{\Theta_1, \Theta_2}
-%%}
-%%\end{mathpar}
-%%$\fbox{$\exprOKX{e}$}$
-%%~~~$\itvarCtx ::= \emptyctx \pipe \itvarCtx, \evar{x}$
-%%\begin{mathpar}
-%%\inferrule[e-var-kinding]{ }{
-%%	\exprOK{\tvarCtx}{\itvarCtx, \evar{x}}{\fvalCtx}{\evar{x}}
-%%}
-%%
-%%\inferrule[e-lam-kinding]{
-%%	\tKindX{\tau}{\kTypeBlur}\\
-%%	\exprOK{\tvarCtx}{\eivarCtxX{x}}{\fvalCtx}{e}
-%%}{
-%%	\exprOKX{\elam{\evar{x}}{\tau}{e}}
-%%}
-%%
-%%\inferrule[e-op-kinding]{
-%%	\fvarOfType{Fam}{\kappaidx}{\Theta} \in \fvalCtx\\
-%%	\kOpS{op}{\kappai} \in \Theta\\
-%%	\tKindX{\taui}{\kappai}\\\\
-%%	\exprOKX{e_1}\\
-%%	\cdots\\
-%%	\exprOKX{e_n}
-%%}{
-%%	\exprOKX{\eop{Fam}{op}{\taui}{\splat{e}{1}{n}}}
-%%}
-%%\end{mathpar}
-%%\caption{\small Kinding for programs. Variable contexts $\tvarCtx$ and $\itvarCtx$ obey standard structural properties. Kinding rules for type-level terms are given in Figure \ref{tlkind}.}
-%%\label{kindprog}
-%%\vspace{-10pt}
-%%\end{figure}
-%%%\subsection{Active Typechecking and Translation}
-%%The second phase of compilation involves invoking the  logic implemented by user-defined operator families to typecheck and translate the program.
-The rules for this phase are given in Fig.
-\ref{att}.
-The context $\fvalCtx$ tracks the operator definitions and representation schemas associated with families in scope.
-Because the logic inside operators must be invoked during this phase, we need an evaluation semantics for type-level terms, given in Fig.
-\ref{tleval}.
-This is again a largely unsurprising collection of rules with the exception of \textsc{dencase-eval-valid}, which we will explain below.
-%%
-%%\subsection{Abstract Representations and Abstract Internal Types}\label{repcon}
-%%To typecheck and translate an expression, $e$, the rule \textsc{att-exp} first assigns a type, $\tau$, and an \emph{abstract translation}, $\gabs$, to it, as determined by the judgement $\ecompilesX{e}{\tau}{\gabs}$.
-It then \emph{deabstracts} this abstract translation to complete the compilation process, as determined by the judgement $\eraseX{\gabs}{\iota}$.
-The purpose of the abstract translation phase is to ensure that the implementation details of type families are not exposed to other families, so that invariants that they rely on are preserved when families are composed.
-For example, our implementation of natural numbers as integers in Fig.
-\ref{example} maintains the invariant that the translation is non-negative.
-If the knowledge that natural numbers are implemented as integers was externally visible, a different extension could introduce an operator like $\tops{badnat}{\kunit}{i}{a}{\tapp{\tvar{const}}{\tapp{\tvar{a}}{\tden{\titerm{-1}}{\ttype{Nat}{\tunit}}}}}$, breaking this invariant (and thus the guarantee that the recursor always terminates!)
-%%%To put it another way, if a collection of operators can be shown to be a full and faithful (that is, adequate) encoding of a particular type system, then it does not matter what other types there are in the system.
-%%
-%%We preclude such operations by a mechanism similar to the abstract type mechanism supported by ML-style module systems \cite{pfpl}.
-By keeping a type family's representation schema private to the operators associated with it, they maintain full control over representation invariants.
-So, because it cannot be shown given the knowledge available in the type family containing $\opvar{badnum}$ that the internal type associated with $\ttype{Nat}{\tunit}$ is $\dint$, the translation $-1$ will not be permitted according to the rules we will give below.
-The only way to produce a term of type $\ttype{Nat}{\tunit}$ as the result of applying an operator not associated with the family $\fvar{Nat}$ is if that operator extracts the term from an argument (e.g.
-when projecting it out of an $n$-tuple), in which case the necessary invariants are inductively  maintained (see Sec.
-\ref{safety}).
-Translations extracted from denotations via \textsf{case} analysis are tracked during this phase as \emph{abstract internal terms}, $\tvalof{\tau_1}{\tau_2}$ (see rule \textsc{dencase-eval-valid}).%
-%%
-%%Let us first review how expressions are assigned types and abstract translations.
-Variables (\textsc{att-var}) translate directly to variables in the internal language, with the type determined by the typing context, $\iota$.
-Lambda terms (\textsc{att-lam}) also translate to lambda terms in the internal language and are given the appropriate $\fvar{Arrow}$ type by extending the typing context and proceeding recursively as is usual.
-The internal type of the argument, however, is left as $\trepof{\tau_1'}$, which is called the \emph{abstract representation} of $\tau_1'$.
-The actual representation of this type is only available from within operators in its family.
-%%
-%%Now we will consider the important operator application rule (\textsc{att-op}).
-This rule operates by extracting the appropriate operator definition, evaluating the operator index to a value, $\tauidx'$, and recursively assigning a type and abstract translation to each argument.
-From these, it constructs a list of denotations  and passes this list, along with the fully evaluated operator index, to the  operator implementation, $\tauop$.
-If this results in a valid denotation, compilation can proceed, but if an $\terr$ is produced, compilation will stop because no other rule will apply (in practice, we would display a type error at this point).
-%%
-%%Before a valid denotation produced in this way can be used, however, we check that it is \emph{representationally consistent}, meaning that the abstract translation, $\gabs$, is of an abstract internal type consistent with the abstraction representation of the type it is paired with, $\tau$.
-The premise $\ddbar{\Xi_0,\fvar{Fam}}{\fvalCtx}{\trepof{\tau}}{\sabs}$ determines the abstract internal type and the premise $\checkRC{\gtCtx}{\Xi_0,\fvar{Fam}}{\fvalCtx}{\gabs}{\sabs}$ performs the internal type checking.
-To do so, the variable context, $\iota$, which associates variables with types, must be converted to a context associating those variables with their corresponding abstract internal types, $\gtCtx$.
-The relevant judgements are defined in Fig.
-\ref{ait}.
-The schema context $\Xi$ is used to track which representation schemas are visible.
-In our calculus, $\Xi_0=\fvar{Arrow}$ is always visible alongside the schema of the family associated with the operator being considered.
-This could be exanded to support module-scoped visibility (we do not include modules in our calculus for simplicity), mutually-defined type families or type families that intentionally expose their representation schemas publicly (as $\fvar{Arrow}$ does).
-%%
-%%To clarify this fundamental mechanism let us examine how the expression $\eop{Nat}{s}{\tunit}{\eop{Nat}{z}{\tunit}{ }}$ will be processed during this phase.
-The inner application of $\opvar{z}$ will produce the abstract translation $0$ paired with the type $\ttype{Nat}{\tunit}$.
-Because the representation schema for $\fvar{Nat}$ is available, we have that $\ddbar{\Xi_0,\fvar{Nat}}{\fvalCtx}{\trepof{\ttype{Nat}{\tunit}}}{\dint}$ by \textsc{show-rep}, as needed.
-In contrast, if $\opvar{badnat}$ were used inside, the representation schema of $\fvar{Nat}$ would not be available, so we can only apply \textsc{hide-rep}, which does not give us enough information to give $-1$ a type.
-%%
-%%When the compiler next considers the outer application of $\opvar{s}$, it will pass the denotation $\tden{\titerm{0}}{\ttype{Nat}{\tunit}}$ into its definition.
-There, it binds the translation to the variable $\tvar{x}$ (within the $\tvar{pop\_final}$ helper function).
-However, $\tvar{x}$ is not simply $\titerm{0}$.
-Instead, its provenance is tracked by using the form $\tvalof{\titerm{0}}{\ttype{Nat}{\tunit}}$.
-When attempting to check that $\iup{\tvar{x}} + 1$ is valid, the \textsc{show-trans} rule will reveal that it is an integer because, again, the appropriate representation schema is available.
-If the schema were not available, the most that could be derived is that $\checkRC{\gtCtx}{\Xi}{\fvalCtx}{\iup{\tvar{x}}}{\trepof{\ttype{Nat}{\tunit}}}$.
-The fact that natural numbers are represented using integers is not derivable (though it is true), so the addition operation would fail to type.
-This fact would, however, be sufficient for implementing families, like $\fvar{Ntuple}$, that do not require knowledge about a type's representation.
-We encourage the reader to derive these judgements for the logic in the $\opvar{rec}$ operator to strengthen their understanding of this fundamental mechanism.
-%%
-%%We cannot leave abstract representations and abstract internal terms in the result of compilation, so a final deabstraction phase erases these, replacing them with their underlying internal types and terms in all contexts.
-Just as with abstract types in modules or existential types, there is no run-time overhead to this mechanism -- programs run at full speed.
-We give the deabstraction rules in the appendix due to their simplicity.
-%%
-%
-%%
-%
-%%
-%%\begin{figure}[t]
-%%\small
-%%$\fbox{\inferrule{}{\concrep{\fvalCtx}{\tau}{\sigma}}}$
-%%\begin{mathpar}
-%%\inferrule[get-rep]{
-%%	\fval{Fam}{\theta}{i}{\tau} \in \fvalCtx\\
-%%	\tEvalX{[\tauidx/\tvar{i}]\tau}{\titype{\sigma}}\\
-%%	\ddbarX{\sigma}{\sconc}
-%%}{
-%%	\concrep{\fvalCtx}{\ttype{Fam}{\tauidx}}{\sconc}
-%%}
-%%\end{mathpar}
-%%$\fbox{\inferrule{}{\ddbarX{\sigma}{\sigma'}}}$
-%%~~~$\fvalCtx ::= \fvalCtx_0 \pipe \fvalCtx, \fvalDf$~~~
-%%~~~$\Xi ::= \Xi_0 \pipe \Xi, \fvar{Fam}$~~~
-%%~~~$\Xi_0 := \fvar{Arrow}$
-%%\begin{mathpar}
-%%\inferrule[abs-int]{ }{
-%%	\ddbarX{\dint}{\dint}
-%%}
-%%
-%%\inferrule[abs-arrow]{
-%%	\ddbarX{\sigma_1}{\sigma_1'}\\
-%%	\ddbarX{\sigma_2}{\sigma_2'}
-%%}{
-%%	\ddbarX{\darrow{\sigma_1}{\sigma_2}}{\darrow{\sigma_1'}{\sigma_2'}}
-%%}
-%%
-%%\inferrule[abs-prod]{
-%%	\ddbarX{\sigma_1}{\sigma_1'}\\
-%%	\ddbarX{\sigma_2}{\sigma_2'}
-%%}{
-%%	\ddbarX{\dpair{\sigma_1}{\sigma_2}}{\dpair{\sigma_1'}{\sigma_2'}}
-%%}
-%%
-%%\inferrule[itype-inverse]{
-%%	\ddbarX{\sigma}{\sabs}
-%%}{
-%%	\ddbarX{\dup{\titype{\sigma}}}{\sabs}
-%%}
-%%
-%%\inferrule[show-rep]{
-%%	\fvar{Fam} \in \Xi\\
-%%	\concrep{\fvalCtx}{\ttype{Fam}{\tauidx}}{\sconc}
-%%}{
-%%	\ddbarX{\trepof{\ttype{Fam}{\tauidx}}}{\sconc}
-%%}
-%%
-%%\inferrule[hide-rep]{
-%%	\fvar{Fam} \notin \Xi
-%%}{
-%%	\ddbarX{\trepof{\ttype{Fam}{\tauidx}}}{\trepof{\ttype{Fam}{\tauidx}}}
-%%}
-%%\end{mathpar}
-%%$\fbox{\inferrule{}{\eCtxTogCtxX{\etCtx}{\gtCtx}}}$
-%%~~~$\gtCtx ::= \emptyctx \pipe \gtCtxX{x}{\sigma}$
-%%\begin{mathpar}
-%%\inferrule[abs-empty]{ }{
-%%	\eCtxTogCtxX{\emptyctx}{\emptyctx}
-%%}
-%%
-%%\inferrule[abs-ctx]{
-%%	\eCtxTogCtxX{\etCtx}{\gtCtx}\\
-%%	\ddbarX{\trepof{\tau}}{\sigma}
-%%}{
-%%	\eCtxTogCtxX{\etCtxX{x}{\tau}}{\gtCtxX{x}{\sigma}}
-%%}
-%%\end{mathpar}
-%%$\fbox{\inferrule{}{\checkRCX{\iota}{\sigma}}}$
-%%\begin{mathpar}
-%%\inferrule[abs-i-var]{ }{
-%%	\checkRC{\gtCtxX{x}{\sigma}}{\Xi}{\fvalCtx}{\evar{x}}{\sigma}
-%%}
-%%
-%%\inferrule[abs-i-lam]{
-%%	\ddbarX{\sigma_1}{\sigma_1'}\\
-%%	\checkRC{\gtCtxX{x}{\sigma_1'}}{\Xi}{\fvalCtx}{\iota}{\sigma_2}
-%%}{
-%%	\checkRCX{\ilam{x}{\sigma_1}{\iota}}{\darrow{\sigma_1'}{\sigma_2}}
-%%}
-%%
-%%\inferrule[abs-i-ap]{
-%%	\checkRCX{\iota_1}{\darrow{\sigma_1}{\sigma_2}}\\
-%%	\checkRCX{\iota_2}{\sigma_1}
-%%}{
-%%	\checkRCX{\iapp{\iota_1}{\iota_2}}{\sigma_2}
-%%}
-%%
-%%\inferrule[abs-i-fix]{
-%%	\ddbarX{\sigma}{\sigma'}\\
-%%	\checkRC{\gtCtxX{x}{\sigma'}}{\Xi}{\fvalCtx}{\iota}{\sigma'}
-%%}{
-%%	\checkRCX{\ifix{x}{\sigma}{\gabs}}{\sigma'}
-%%}
-%%\\
-%%\text{\color{gray} (standard statics for integers and products omitted)}
-%%\\
-%%%\inferrule[abs int]{ }{
-%%%	\checkRCX{\iintlit}{\dint}
-%%%}
-%%%
-%%%\inferrule[abs op]{
-%%%	\checkRCX{\iota_1}{\dint}\\
-%%%	\checkRCX{\iota_2}{\dint}
-%%%}{
-%%%	\checkRCX{\iop{\iota_1}{\iota_2}}{\dint}
-%%%}
-%%%
-%%%\inferrule[abs pair]{
-%%%	\checkRCX{\iota_1}{\sigma_1}\\
-%%%	\checkRCX{\iota_2}{\sigma_2}
-%%%}{
-%%%	\checkRCX{\ipair{\iota_1}{\iota_2}}{\dpair{\sigma_1}{\sigma_2}}
-%%%}
-%%%
-%%%\inferrule[abs fst]{
-%%%	\checkRCX{\iota}{\dpair{\sigma_1}{\sigma_2}}
-%%%}{
-%%%	\checkRCX{\ifst{\iota}}{\sigma_1}
-%%%}
-%%%
-%%%\inferrule[abs snd]{
-%%%	\checkRCX{\iota}{\dpair{\sigma_1}{\sigma_2}}
-%%%}{
-%%%	\checkRCX{\isnd{\iota}}{\sigma_2}
-%%%}
-%%%
-%%% \inferrule[abs-int-eq]{
-%%% 	\checkRCX{\iota_1}{\dint}\\
-%%% 	\checkRCX{\iota_2}{\dint}\\\\
-%%% 	\checkRCX{\iota_3}{\sigma}\\
-%%% 	\checkRCX{\iota_4}{\sigma}
-%%% }{
-%%% 	\checkRCX{\iIfEq{\iota_1}{\iota_2}{\dint}{\iota_3}{\iota_4}}{\sigma}
-%%% }
-%%\inferrule[abs-iterm-inverse]{
-%%	\checkRCX{\iota}{\sigma}
-%%}{
-%%	\checkRCX{\iup{\titerm{\iota}}}{\sigma}
-%%}
-%%
-%%\inferrule[show-trans]{
-%%	\fvar{Fam} \in \Xi\\
-%%	\concrep{\fvalCtx}{\ttype{Fam}{\tauidx}}{\sigma}\\
-%%	\checkRCX{\iota}{\sigma}
-%%}{
-%%	\checkRCX{\tvalof{\titerm{\iota}}{\ttype{Fam}{\tauidx}}}{\sigma}
-%%}
-%%
-%%\inferrule[hide-trans]{
-%%	\fvar{Fam} \notin \Xi\\
-%%	\concrep{\fvalCtx}{\ttype{Fam}{\tauidx}}{\sigma}\\
-%%	\checkRCX{\iota}{\sigma}
-%%}{
-%%	\checkRCX{\tvalof{\titerm{\iota}}{\ttype{Fam}{\tauidx}}}{\trepof{\ttype{Fam}{\tauidx}}}
-%%}
-%%\end{mathpar}
-%%\caption{\small Abstracted internal typing}
-%%\label{ait}
-%%\end{figure}
-%%%
-%%%family RECORD of (string, type) dict {
-%%%  new(idx, args. foldl (fn (x, y) => 
-%%%  get[field : string](idx. case find(idx, field) of SOME t => t | _ => err)
-%%
-%%\section{Safety of @$\lambda$}\label{safety}
-%%In giving users of a language direct influence on the typechecking and translation of expressions, it is essential to consider safety properties.
-It must not be possible for well-typed programs to fail to finish compiling or go wrong at run-time because of a buggy extension.
-It is also considered desirable for typechecking to be decidable.
-And for extensions to be reliable, they must not be allowed to interfere with one another under any circumstance.
-By carefully designing our extension mechanism, we believe we have achieved each of these goals.
-%%
-%%\subsection{Representational Consistency and Type Safety}
-%%We do not need to give a semantics to internal terms and internal types that do not survive deabstraction.
-The remaining terms and types form a variant of PCF with well-understood primitives (here, integers and binary products) known to be type safe \cite{pfpl}.
-If compilation can be shown to always result in a well-typed term of this language, then type safety follows by composing these two facts.
-Fortunately, this fact is a corollary of our representational consistency mechanism, described in Sec.
-\ref{repcon}.
-The proof is by straightforward (though deeply nested!) rule induction.
-%%
-%%%\begin{theorem}[Representational Consistency]
-%%%If $\progOK{ }{\Sigma}{e}$ and $\ecompiles{ }{\fvalCtx}{e}{\tau}{\gabs}$ and $\eraseX{\gabs}{\iota}$ and 	$\ddbar{\Xi_0}{\fvalCtx}{\trepof{\tau}}{\sabs}$ and $\eraseX{\sabs}{\sigma}$ then $\checkRC{ }{\Xi_0}{\fvalCtx}{\iota}{\sigma}$.
-%%%\end{theorem}
-%%
-%%\subsection{Decidability}
-%%To show that compilation is decidable, we need to show that both kind checking and active typechecking and compilation are decidable.
-This hinges on showing that the type-level language is typesafe and that evaluation of type-level terms always terminates.
-%%
-%%%\begin{theorem}[Kind Safety and Termination]
-%%%If $\vdash_{\Sigma} \tau : \kappa$ then $\tEvalX{\tau}{\tau'}$ such that $\vdash_{\Sigma_0} \tau' : \kappa$ and $\tEvalX{\tau'}{\tau'}$. 
-%%%\end{theorem}
-%%
-%%The proof is straightforward because the type-level language is based on a conventional simply-typed lambda calculus with only bounded recursion over lists, so standard techniques (e.g.
-logical relations) can be used directly.
-The only new constructs are the types, denotations and reified internal language forms, but because none of these can contain type-level functions by the statics, there is no risk of introducing self-reference by their inclusion.
-If the type-level language included constructs like inductive datatypes (without a strict positivity condition) or general recursion, this theorem would be weakened.
-%%
-%%\subsection{Non-Interference}
-%%Our abstraction mechanism guarantees that the representation invariants collectively maintained by the operators associated with a type family cannot be violated by operators in other type families, by ensuring that introductory forms cannot be defined outside the family.
-One way to state this is as follows:
-%%
-%%%\begin{theorem}[Non-Interference]
-%%%For any property $P(\iota)$, if $\progOK{ }{\Sigma}{e}$ and $\ecompiles{ }{\fvalCtx}{e}{\tau}{\gabs}$ and $\eraseX{\gabs}{\iota}$ implies $P(\iota)$, then for any $\Sigma'$ and $\fvalCtx'$ disjoint from $\Sigma$ and $\fvalCtx$, we have that if $\progOK{ }{\Sigma,\Sigma'}{e'}$ and $\ecompiles{ }{\fvalCtx,\fvalCtx'}{e'}{\tau}{\gabs'}$ and $\eraseX{\gabs'}{\iota'}$ then $P(\iota')$.
-%%%\end{theorem}
-%%The proof relies on the fact that the show/hide rules are mutually exclusive and so there are two cases to consider for each form of expression, and weakening properties of the family contexts.
-This powerful theorem allows us to compose type families arbitrarily without needing to handle cases in our implementation that are ruled out based on a local analysis.
-The details will be provided in the appendix.
-%%
-%%% \subsection{Modular Verification}
-%%% nope
-%%
-%%% that's all folks
-%%
-%%\section{Use Cases}\label{examples}
-%%
-%%Aldrich argues that dynamic dispatch has proven useful for building  large software frameworks \cite{aldrich-onward13}.
-Object-oriented languages support dynamic dispatch directly but encodings of dynamic dispatch in terms of standard functional primitives can be awkward and difficult to reason about (e.g.
-\cite{pfpl} Ch.
-X).
-%%
-%%Indexed type families and indexed operator families are a powerful tool for programming language semantics.
-In \emph{Practical Foundations for Programming Languages}, for example, nearly every chapter defines a new collection of indexed types and operators and gives their semantics in isolation from the constructs in the other chapters \cite{pfpl}.
-For example, $n$-ary sums and products are families indexed by a list of types.
-Labeled variants of these simply pair the labels with the types as static strings.
-Even nested pattern matching, such as that found in modern functional languages, can be understood as an operator family indexed by a series of patterns, which can be represented as type-level data.
-In the appendix, we show how to implement sums and products in \atlam.
-%%
-%%The book also shows constructs as varied as dynamic types as well as a number of constructs for parallelism and concurrency, such as futures, promises and actors.
-With a sufficiently capable internal language (exposing basic concurrency primitives, for example), each of these could be fully implemented as libraries using our mechanism.
-Few languages include primitive support for several such abstractions, even though they are often useful together.
-In a preliminary implementation of this calculus, Ace, which is beyond the scope of this paper, we have implemented the entirety of the OpenCL programming language as a library, along with primitives supporting partitioned global address spaces.
-%%
-%%Object systems too could be implemented using this mechanism, with indices serving to capture the inheritance data and the signatures.
-Operator families for reading and writing fields and/or sending messages would be parameterized by strings naming them.
-The operator definition would simply search the signatures going up the inheritance hierarchy, and implement objects in a conventional way (e.g.
-using a v-table together with a pointer to a structure containing field data).
-A variety of object systems could coexist within the same language.
-Another related example would be to implement a safe and efficient interoperability layer with an existing OO language by capturing its type system as an extension, to be used only at language boundaries.
-%%
-%%Finally, a variety of domain-specific type systems, capturing complex rule systems like the system of scientific units of measure (a family indexed by the measure being used and the type which the measure is being applied to), regular expressions \cite{regexp} (a family indexed by the number of captured groups) or XML \cite{xml} (indexed by a document schema) would be possible.
-Recent work examined a specialized type system capturing the semantics of the widely-used jQuery javascript library \cite{jquery-oopsla}.
-Today, these all require \emph{ad hoc} language-external solutions, or the development of new languages.
-%%
-%%%\section{Design Considerations}\label{design}
-%%
-%\section{Related Work}\label{related-work}
-%Our representation schema abstraction mechanism relates closely to abstract and existential types \cite{pfpl,atpl}.
-Our calculus enforces the abstraction barriers  in a purely syntactic manner, as in previous work on syntactic type abstraction \cite{syntypeabs}.
-While this work is all focused on abstracting away the identity of a particular type outside of the ``principal'' it is associated with (e.g.
-a module), we focus on abstracting away the knowledge of how a primitive type family is implemented outside of a limited scope.
-%
-%The representational consistency mechanism brings into the language work on typed compilation, especially work done on Standard ML in the TILT project \cite{tilt}.
-Indeed, the specification of Standard ML is structured around a typed internal language and a judgement that assigns a type and an internal term to each expression \cite{smlstd}.
-Representational consistency is related to the notion of type-directed copmilation in this work.
-%
-%Type-level computation is supported in some form by a growing number of languages.
-For example, Haskell supports a simple form of it \cite{Chakravarty:2005:ATC}).
-Ur uses type-level records and names to support typesafe metaprogramming, with applications to web programming \cite{conf/pldi/Chlipala10}.
-$\Omega$mega adds algebraic data types at the type-level, using these to increase the expressive power of algebraic data types at the expression level \cite{conf/cefp/SheardL07}.
-Dependently-typed languages blur the traditional phase separation between types and expressions, so type-level computation is often implicitly used (though not always in its most general form, e.g.
-Deputy \cite{conf/icfp/ChenX05}, ATS \cite{conf/esop/ConditHAGN07}).
-We show how to integrate language extensions into the type-level language, drawing on ideas about \cite{activelibraries}.
-%
-%% \subsection{Run-Time Indirection}
-%% {\it Operator overloading} \cite{vanWijngaarden:Mailloux:Peck:Koster:Sintzoff:Lindsey:Meertens:Fisker:acta:1975} and {\it metaobject dispatch} \cite{Kiczales91} are run-time protocols that translate operator invocations into function calls.
-The function is typically selected according to the type or value of one or more operands.
-These protocols share the notion of {\it inversion of control} with type-level specification.
-However, type-level specification is a {\it compile-time} protocol focused on enabling specialized verification and implementation strategies, rather than simply enabling run-time indirection.
-%
-%Many languages and tools allow developers to rewrite expressions according to custom rules.
-These can broadly be classified as {\it term rewriting systems}.
-Macro systems, such as those characteristic of the LISP family of languages \cite{mccarthy1978history}, are the most prominent example.
-Some compile-time metaprogramming systems also allow users to manipulate syntax trees (e.g.
-MetaML \cite{Sheard:1999:UMS}), and external rewrite systems also exist for many languages.
-%These facilities differ from AT\&T in that they involve direct manipulation of terms, while AT\&T involves extending typechecking and translation logic directly.
-We also draw a distinction between the type-level, used to specify types and compile-time logic, the expression grammar, used to describe run-time behavior, and the internal language, used to implement this behavior.
-By doing so, each component can be structured and constrained as appropriate for its distinct role, as we have shown.
-%
-%Previous work on extensible languages has suffered from problems with either expressiveness or safety.
-For example, a number of projects, such as SugarJ \cite{erdweg2011sugarj}, allow for user-defined desugarings (and indeed, our system would clearly benefit from integration with such a mechanism), but this does not allow the semantics to be fundamentally extended nor for implementation details to be hidden.
-Recent variants of this work has investigated introducing new typing rules, but only if they are admissible by the base type system \cite{erdwegicfp}.
-Our work allows for entirely new logic to be added to the system, requiring only that the implementation of this logic respect the internal type system.
-A number of other extensible languages (e.g.
-Xroma \cite{xroma}) and compilers do allow new static semantics to be introduced, but in an unconstrained manner based on global pattern matching and rewriting, leading to significant problems with interference and safety.
-Our work aims to provide a sound and safe underpinning, based around well-understood concepts of type and operator families, to language extensibility.
-%
-%In the future, we will investigate mechanisms to enable type systems with specialized binding and scoping rules, as well as integration of dependent kinds to support mechanized verification of key properties like representational consistency and adequacy against a declarative specification at kind-checking time.
-We will also investigate mechanisms that enable a more natural syntax (e.g.
-Wyvern's type-directed syntax \cite{globaldsl}).
-%
-%%\item With type-level specification, dispatch to a type-level function occurs implicitly on the basis of the structure of an expression.
-In contrast, most term-rewriting systems operate by  explicit invocation of a macro or specialized syntax.
-Some LISP macro systems have explored pattern-based dispatch (e.g.
-A*\cite{fowler2010domain}, EPP\cite{fowler2010domain}) and macro systems for object-oriented languages, like OpenC++ \cite{fowler2010domain} and OpenJava \cite{fowler2010domain}, do offer a somewhat limited form of operation-based dispatch.
-%
-%% \subsection{Language Frameworks}
-%% When the mechanisms available in an existing language prove insufficient, researchers and domain experts must design a new language.
-A number of tools have been developed to assist with this task, including compiler generators, language workbenches and domain-specific language frameworks (cf \cite{fowler2010domain}).
-%
-%% A major barrier to adoption is the fact that interoperability is intrinsically problematic.
-Even languages which target a common platform, such as the Java Virtual Machine, can only interact using its limited set of primitives.
-Specialized typing rules are not checked at language boundaries, performance often suffers, and the syntax can be unnatural, particularly for languages which differ significantly from the platform's native language (e.g.
-Java).
-%
-%% Instead of focusing on defining standalone languages, type-level specification gives greater responsibility in a granular manner to libraries.
-In this way, a range of constructs can coexist within the same program and, assuming that it can be shown by some method that various constructs are safely composable, be mixed and matched.
-The main limitation is that the protocol requires defining a fixed source grammar, whereas a specialized language has considerable flexibility in that regard.
-Nevertheless, as Ace shows, a simple grammar can be used quite flexibly.
-%% \subsection{Extensible Compilers}
-%% An alternative methodology is to implement language features granularly as compiler extensions.
-As discussed in Section 1, existing designs suffer from the same problems related to composability, modularity\-, safety and security as extensible languages, while also adding the issue of language fragmentation.
-%
-%% Type-level specification can in fact be implemented within a compiler, rather than provided as a core language feature.
-This would resolve some of the issues, as described in this paper.
-However, by leveraging type-level computation to integrate the protocol directly into the language, we benefit from common module systems and other shared infrastructure.
-We also avoid the fragmentation issue.
-%% \subsection{Specification Languages}
-%% Several {\it specification languages} (or {\it logical frameworks}) based on these theoretical formulations exist, including the OBJ family of languages (e.g.
-CafeOBJ \cite{Diaconescu-Futatsugi01}).
-They provide support for verifying a program against a language specification, and can automatically execute these programs as well in some cases.
-The  language itself specifies which verification and execution strategies are used.
-%
-%% Type-determined compilation takes a more concrete approach to the problem, focusing on combining {\it implementations} of different\- logics, rather than simply their specifications.
-In other words, it focuses on combining {\it type checkers} and {\it implementation strategies} rather than more abstract representations of a language's type system and dynamic semantics.
-In Section 4, we outlined a preliminary approach based on proof assistant available for the type-level language to unify these approaches, and we hope to continue this line of research in future work.
-%
-%% CANT GUARANTEE THAT SPECIFICATIONS ARE ACTUALLY DECIDABLE
-%\begin{figure}
-%\small
-%$\fbox{$\tEvalX{\tau}{\tau'}$}$
-%\begin{mathpar}
-%\inferrule[n-t-lam]{ }{
-%	\tEvalX{\tlam{t}{\kappa}{\tau}}{\tlam{t}{\kappa}{\tau}}
-%}
-%
-%\inferrule[n-t-ap]{
-%	\tEvalX{\tau_1}{\tlam{t}{\kappa}{\tau}}\\
-%	\tEvalX{\tau_2}{\tau_2'}\\
-%	\tEvalX{\subst{\tau_2'}{\tvar{t}}{\tau}}{\tau'}
-%}{
-%	\tEvalX{\tapp{\tau_1}{\tau_2}}{\tau'}
-%}
-%
-%\text{\color{gray} (normalization for integers, labels, lists, products and sums also standard)}
-%
-%\inferrule[n-t-eq-equal]{
-%	\tEvalX{\tau_1}{\tau_1'}\\
-%	\tEvalX{\tau_2}{\tau_1'}\\
-%	\tEvalX{\tau_3}{\tau_3'}
-%}{
-%	\tEvalX{\tifeq{\tau_1}{\tau_2}{\kappa}{\tau_3}{\tau_4}}{\tau_3'}
-%}
-%
-%\inferrule[n-t-eq-disequal]{
-%	\tEvalX{\tau_1}{\tau_1'}\\
-%	\tEvalX{\tau_2}{\tau_2'}\\\\
-%	\tau_1' \neq \tau_2'\\
-%	\tEvalX{\tau_4}{\tau_4'}
-%}{
-%	\tEvalX{\tifeq{\tau_1}{\tau_2}{\kappa}{\tau_3}{\tau_4}}{\tau_4'}
-%}
-%
-%\inferrule[nt-ty]{
-%	\tEvalX{\tauidx}{\tauidx'}
-%}{
-%	\tEvalX{\ttype{Tycon}{\tauidx}}{\ttype{Tycon}{\tauidx'}}
-%}
-%
-%\inferrule[nt-tycase-match]{
-%	\tEvalX{\tau}{\ttype{Tycon}{\tauidx}}\\
-%	\tEvalX{\subst{\tauidx}{\tvar{x}}{\tau_1}}{\tau_1'}
-%}{
-%	\tEvalX{\tfamcase{\tau}{Tycon}{x}{\tau_1}{\tau_2}}{\tau_1'}
-%}
-%
-%\inferrule[nt-tycase-fail]{
-%	\tEvalX{\tau}{\ttype{Tycon'}{\tauidx}}\\
-%	\fvar{Tycon} \neq \fvar{Tycon'}\\
-%	\tEvalX{\tau_2}{\tau_2'}
-%}{
-%	\tEvalX{\tfamcase{\tau}{Tycon}{x}{\tau_1}{\tau_2}}{\tau_2'}
-%}
-%
-%\inferrule[nt-d]{
-%	\tEvalX{\tau}{\tau'}\\
-%	\tiEvalX{\ibar}{\ibar'}
-%}{
-%	\tEvalX{\tden{\ibar}{\tau}}{\tden{\ibar'}{\tau'}}
-%}
-%
-%\inferrule[nt-d-checked]{
-%	\tEvalX{\tau}{\tau'}\\
-%	\tiEvalX{\ibar}{\ibar'}
-%}{
-%	\tEvalX{\tden{\ibar}{\tau}^\checkmark}{\tden{\ibar'}{\tau'}^\checkmark}
-%}
-%
-%\inferrule[nt-typeof]{
-%	\tEvalX{\tau}{\tden{\ibar}{\tau'}}
-%}{
-%	\tEvalX{\ttypeof{\tau}}{\tau'}
-%}
-%
-%\inferrule[nt-typeof-checked]{
-%	\tEvalX{\tau}{\tden{\ibar}{\tau'}^\checkmark}
-%}{
-%	\tEvalX{\ttypeof{\tau}}{\tau'}
-%}
-%
-%\inferrule[nt-reptype]{
-%	\tiEvalX{\sbar}{\sbar'}
-%}{
-%	\tEvalX{\titype{\sbar}}{\titype{\sbar'}}
-%}
-%\end{mathpar}
-%$\fbox{$\tiEvalX{\ibar}{\ibar'}$}$
-%\begin{mathpar}
-%\inferrule[nt-i-var]{ }{
-%	\tiEvalX{\evar{x}}{\evar{x}}
-%}
-%~~~~~
-%%\inferrule[nt-i-lam]{
-%%	\tiEvalX{\sigma}{\sigma'}\\
-%%	\tiEvalX{\iota}{\iota'}
-%%}{
-%%	\tiEvalX{\ilam{\evar{x}}{\sigma}{\iota}}{\ilam{\evar{x}}{\sigma'}{\iota'}}
-%%}
-%%
-%\inferrule[nt-i-fix]{
-%	\tiEvalX{\sbar}{\sbar'}\\
-%	\tiEvalX{\ibar}{\ibar'}
-%}{
-%	\tiEvalX{\ifix{\evar{x}}{\sbar}{\ibar}}{\ifix{\evar{x}}{\sbar'}{\ibar'}}
-%}
-%~~~~~
-%\inferrule[nt-i-transof]{
-%	\tEvalX{\tau}{\tau'}
-%}{
-%	\tiEvalX{\itransof{\tau}}{\itransof{\tau'}}
-%}
-%\\
-%\text{\color{gray} (omitted forms have similarly recursive rules)}
-%%\inferrule[iterm unquote eval]{
-%%	\tEvalX{\tau}{\tau'}
-%%}{
-%%	\tiEvalX{\iup{\tau}}{\iup{\tau'}}
-%%}
-%%
-%\end{mathpar}
-%$\fbox{$\tiEvalX{\sbar}{\sbar'}$}$
-%\begin{mathpar}
-%%\inferrule[i-int-eval]{ }{
-%%	\tiEvalX{\dint}{\dint}
-%%}
-%%
-%%\inferrule[i-prod-eval]{
-%%	\tiEvalX{\sigma_1}{\sigma_1'}\\
-%%	\tiEvalX{\sigma_2}{\sigma_2'}
-%%}{
-%%	\tiEvalX{\dpair{\sigma_1}{\sigma_2}}{\dpair{\sigma_1'}{\sigma_2'}}
-%%}
-%%
-%\inferrule[nt-i-parr]{
-%	\tiEvalX{\sbar_1}{\sbar_1'}\\
-%	\tiEvalX{\sbar_2}{\sbar_2'}
-%}{
-%	\tiEvalX{\darrow{\sbar_1}{\sbar_2}}{\darrow{\sbar_1'}{\sbar_2'}}
-%}
-%
-%\inferrule[nt-i-unquote]{
-%	\tEvalX{\tau}{\tau'}
-%}{
-%	\tiEvalX{\dup{\tau}}{\dup{\tau'}}
-%}
-%
-%\inferrule[nt-i-repof]{
-%	\tEvalX{\tau}{\tau'}
-%}{
-%	\tiEvalX{\trepof{\tau}}{\trepof{\tau'}}
-%}
-%
-%\text{\color{gray} (omitted forms have similarly recursive rules)}
-%\end{mathpar}
-%\caption{\small Normalization semantics for type-level terms, including those inside translational IL. There is nothing clever happening here.}
-%\label{tleval}
-%\end{figure}
-
+% CANT GUARANTEE THAT SPECIFICATIONS ARE ACTUALLY DECIDABLE 
 \bibliographystyle{abbrv}
 \bibliography{../research}
-
-\begin{figure*}[p!]
-\small
-{\color{gray}\text{(constructs lifted from $\mathcal{L}\{{\rightarrow}\,{\forall}\,{\mu_\text{ind}}\,{\kunit}\,{\times}\,{+}\}$ are standard, omitted \cite{pfpl})}}~\hfill\fbox{$\sofk{\kDelta}{\kGamma}{\kXi}{\st}{\kappa\moutput}$}
-~\fbox{$\sqofkX{\qcolor{\ity}}$}~\fbox{$\sqofkX{\qcolor{\itm}}$}
-\begin{mathpar}
-\inferrule[k-ty]{
-  c[\ktyidx] \in \kXi\\\\
-  \sofkX{\sttyidx}{\ktyidx}
-}{
-  \sofkX{\sty{c}{\sttyidx}}{\kty}
-}
-
-\inferrule[k-otherty]{ }{
-  \sofkX{\sotherty{m}{\tau}}{\kty}
-}
-
-\inferrule[k-tycase]{
-  c[\ktyidx] \in \kXi\\\\
-  \sofkX{\st}{\kty}\\
-  \sofk{\kDelta}{\kGamma, \sx : \ktyidx}{\kXi}{\st_1}{\kappa}\\
-  \sofkX{\st_2}{\kappa}
-}{
-  \sofkX{\stycase{c}{\st}{\sx}{\st_1}{\st_2}}{\kappa}
-}
-
-\inferrule[k-qity]{\sqofkX{\qcolor{\ity}}}{
-  \sofkX{\sqity{\qcolor{\ity}}}{\kity}
-}
-
-\inferrule[k-qity-uq]{
-  \sofkX{\st}{\kity}
-}{
-  \sqofkX{\qcolorc\qtuq{\color{black}\st\qcolorc}\color{black}}
-}
-
-\inferrule[k-rep]{
-  \sofkX{\st}{\kty}
-}{
-  \sofkX{\srep{\st}}{\kity}
-}
-
-%\inferrule[kity]{a}{b}
-%
-\inferrule[k-qitm]{
-  \sqofkX{\qcolor{\itm}}
-}{
-  \sofkX{\sqitm{\qcolor{\itm}}}{\kitm}
-}
-
-\inferrule[k-qitm-uq]{
-  \sofkX{\st}{\kitm}
-}{
-  \sqofkX{\qcolorc\quq{\color{black}\st\qcolorc}\color{black}}
-}
-
-\text{\color{gray} (rules for remaining \qcolor{quoted forms} generically recursive)}
-
-\inferrule[k-syn]{
-  n' < n
-}{
-  \sofkX{\ssyn{n'}}{\kty \times \kitm}
-}
-
-%\inferrule[kitm]{a}{b}
-%
-\inferrule[k-ana]{
-  n' < n\\
-  \sofkX{\st}{\kty}
-}{
-  \sofkX{\sana{n'}{\st}}{\kitm}
-}
-
-\inferrule[k-raise]{
-  \kok{\kDelta}{\emptyset}{\kappa}
-}{
-  \sofkX{\sraise{\kappa}}{\kappa}
-}
-
-\end{mathpar}
-\caption{Kinding for static terms.}
-\label{statics-SL}
-\end{figure*}
-\begin{figure*}[p!]
-\small
-$\begin{array}{llll}
-\mvlbl{concrete rep whitelist} & \mvlbl{abstract rep store} & \mvlbl{argument store} & \mvlbl{operation context}\\
-\boxdot ::= {\rightharpoonup} ~|~ \boxdot, \tc \ & \memD ::= \emptyset ~|~ \memD, \st \leftrightsquigarrow \alpha & \memG ::= \cdot ~|~ \memG, n \hookrightarrow e ~|~ \memG, n \hookrightarrow e : \st \leadsto \iota/x : \tau & \aCtx ::= \boxdot; \Upsilon; \Phi
-\end{array}$~\hfill{\fbox{$\snorm{\st}{\memD}{\memG}{\aCtx}{\st\moutput}{\memD\moutput}{\memG\moutput}$}}\\
-{\color{gray} (constructs lifted from $\mathcal{L}\{{\rightarrow}\,{\forall}\,{\mu_\text{ind}}\,{\kunit}\,{\times}\,{+}\}$ standard, stores propagate as suggested by rules below, error propagation rules omitted \cite{pfpl})}
-~\hfill\fbox{$\serrX{\st}$}
-\begin{mathpar}
-\inferrule[n-ty]{
-  \snormX{\st}{\st'}
-}{
-  \snormX{\sty{c}{\st}}{\sty{c}{\st'}}
-}
-
-\inferrule[n-otherty]{ }{
-  \snorm{\sotherty{m}{\tau}}{\memD}{\memG}{\aCtx}{\sotherty{m}{\tau}}{\memD}{\memG}
-}
-
-\inferrule[n-tycase-1]{
-  \snorm{\st}{\memD}{\memG}{\aCtx}{\sty{c}{\sttyidx}}{\memD'}{\memG'}\\\\
-  \snorm{[\sttyidx/\svar{x}]\st_1}{\memD'}{\memG'}{\aCtx}{\st_1'}{\memD''}{\memG''}
-}{
-  \snorm{\stycase{c}{\st}{\svar{x}}{\st_1}{\st_2}}{\memD}{\memG}{\aCtx}{\st_1'}{\memD''}{\memG''}
-}
-
-\inferrule[n-tycase-2]{
-  \snorm{\st}{\memD}{\memG}{\aCtx}{\st'}{\memD'}{\memG'}\\\\
-  \st' \neq \sty{c}{\sttyidx}\\
-  \snorm{\st_2}{\memD'}{\memG'}{\aCtx}{\st_2'}{\memD''}{\memG''}
-}{
-  \snorm{\stycase{c}{\st}{\svar{x}}{\st_1}{\st_2}}{\memD}{\memG}{\aCtx}{\st_2'}{\memD''}{\memG''}
-}
-
-\inferrule[n-qity-parr]{
-  \snorm{\sqity{\qcolor{\tau_1}}}{\memD}{\memG}{\aCtx}{\sqity{\tau_1'}}{\memD'}{\memG'}\\\\
-  \snorm{\sqity{\qcolor{\tau_2}}}{\memD'}{\memG'}{\aCtx}{\sqity{\tau_2'}}{\memD''}{\memG''}
-}{
-  \snorm{\sqity{\qcolor{\tau_1 \rightharpoonup \tau_2}}}{\memD}{\memG}{\aCtx}{\sqity{\tau_1' \rightharpoonup \tau_2'}}{\memD''}{\memG''}
-}
-
-\inferrule[n-qity-alpha]{ }{
-  \snorm{\sqity{\qcolor{\alpha}}}{\memD}{\memG}{\aCtx}{\sqity{\alpha}}{\memD}{\memG}
-}
-
-\inferrule[n-qity-forall]{
-  \snormX{\sqity{\qcolor{\tau}}}{\sqity{\tau'}}
-}{
-  \snormX{\sqity{\qcolor{\iforall{\alpha}{\tau}}}}{\sqity{\iforall{\alpha}{\tau'}}}
-}
-
-\text{\color{gray}(other \qcolor{quoted internal types} analagously recursive)}
-
-\inferrule[n-qity-uq]{
-  \snormX{\st}{\sqity{\tau}}
-}{
-  \snormX{\sqity{\qcolorc\qtuq{\color{black}\st\qcolorc}\color{black}}}{\sqity{\tau}}
-}
-
-\inferrule[n-rep-conc-tc]{
-  \snorm{\st}{\memD}{\memG}{\boxdot; \Upsilon; \Phi}{\sty{\tc}{\sttyidx}}{\memD'}{\memG'}\\\\
-  \sty{\tc}{\sttyidx} \leftrightsquigarrow \_ \notin \memD'\\
-  \tc \in \boxdot\\
-  \tcdef{\tc}{\psi}{\omega} \in \Phi\\
-  \keyw{rep}=\stx{rep} \in \omega\\\\
-  \snorm{\stx{rep}~\sttyidx}{\memD'}{\cdot}{\boxdot; \emptyset; \Phi}{\sqity{\tau}}{\memD''}{\cdot}
-}{
-  \snorm{\srep{\st}}{\memD}{\memG}{\boxdot; \Upsilon; \Phi}{\sqity{\tau}}{\memD''}{\memG'}
-}
-
-\inferrule[n-rep-conc-parr]{
-   \snormX{\st}{\sparr{(\st_1, \st_2)}}\\\\
-   \snorm{\srep{\st_1}}{\memD'}{\cdot}{\aCtx}{\sqity{\tau_1}}{\memD''}{\cdot}\\\\
-   \snorm{\srep{\st_2}}{\memD''}{\cdot}{\aCtx}{\sqity{\tau_1}}{\memD'''}{\cdot}
-}{
-   \snorm{\srep{\st}}{\memD}{\memG}{\aCtx}{\sqity{\tau_1 \rightharpoonup \tau_2}}{\memD'''}{\memG'}
-}
-
-\inferrule[n-rep-abs]{
-  \snorm{\st}{\memD}{\memG}{\boxdot; \Upsilon; \Phi}{\sty{c}{\sttyidx}}{\memD'}{\memG'}\\\\
-  \sty{c}{\sttyidx} \leftrightsquigarrow \_ \notin \memD'\\
-  c \notin \boxdot\\
-  (\alpha~\text{fresh})
-}{
-  \snorm{\srep{\st}}{\memD}{\memG}{\boxdot; \Upsilon; \Phi}{\sqity{\alpha}}{\memD', \sty{c}{\sttyidx} \leftrightsquigarrow \alpha}{\memG'}
-}
-
-\inferrule[n-rep-abs-other]{
-  \snormX{\st}{\sotherty{m}{\tau}}\\\\
-  \sotherty{m}{\tau} \leftrightsquigarrow \_ \notin \memD'\\
-  (\alpha~\text{fresh})
-}{
-  \snorm{\srep{\st}}{\memD}{\memG}{\aCtx}{\sqity{\alpha}}{\memD', \sotherty{m}{\tau} \leftrightsquigarrow \alpha}{\memG'}
-}
-
-\inferrule[n-rep-seen]{
-  \snormX{\st}{\st'}\\
-  \st' \leftrightsquigarrow \alpha \in \memD'
-}{
-  \snormX{\srep{\st}}{\sqity{\alpha}}
-}
-
-\inferrule[n-qitm-x]{ }{
-  \snorm{\sqitm{\qcolor{x}}}{\memD}{\memG}{\aCtx}{\sqitm{x}}{\memD}{\memG}
-}
-
-\inferrule[n-qitm-lam]{
-  \snorm{\sqity{\qcolor{\tau}}}{\memD}{\memG}{\aCtx}{\sqity{\tau'}}{\memD'}{\memG'}\\\\
-  \snorm{\sqitm{\qcolor{\iota}}}{\memD'}{\memG'}{\aCtx}{\sqitm{\iota'}}{\memD''}{\memG''}
-}{
-  \snorm{\sqitm{\qcolor{\lambda[\tau](x.\iota)}}}{\memD}{\memG}{\aCtx}{\sqitm{\lambda(x.\iota')}}{\memD''}{\memG''}
-}
-% \inferrule[n-qitm-ap]{
-%   \snorm{\sqitm{\qcolor{\iota_1}}}{\memD}{\memG}{\aCtx}{\sqitm{\iota_1'}}{\memD'}{\memG'}\\\\
-%   \snorm{\sqitm{\qcolor{\iota_2}}}{\memD'}{\memG'}{\aCtx}{\sqitm{\iota_2'}}{\memD''}{\memG''}
-% }{
-%   \snorm{\sqitm{\qcolor{\iap{\iota_1}{\iota_2}}}}{\memD}{\memG}{\aCtx}{\sqitm{\iap{\iota_1'}{\iota_2'}}}{\memD''}{\memG''}
-% }
-
-\text{\color{gray} (other \qcolor{quoted internal terms} analagously recursive)}
-
-\inferrule[n-qitm-uq]{
-  \snormX{\st}{\sqitm{\iota}}
-}{
-  \snormX{\sqitm{\qcolorc\quq{\color{black}\st\qcolorc}\color{black}}}{\sqitm{\iota}}
-}
-
-\inferrule[n-syn]{
-  n \hookrightarrow e \in \memG\\
-  \esyn{\Upsilon}{\Phi}{e}{\st}{\iota}\\
-  (x~\text{fresh})\\\\
-  \snorm{\srep{\st}}{\memD}{\cdot}{\boxdot;\Upsilon;\Phi}{\sqity{\tau}}{\memD'}{\cdot}
-}{
-  \snorm{\ssyn{n}}{\memD}{\memG}{\boxdot; \Upsilon; \Phi}{(\st, \sqitm{x})}{\memD'}{\memG\otimes n \hookrightarrow e : \st \leadsto \iota/x : \tau}
-}
-
-\inferrule[n-syn-seen]{
-  n \hookrightarrow e : \st \leadsto \iota/x : \tau\in \memG
-}{
-  \snorm{\ssyn{n}}{\memD}{\memG}{\aCtx}{(\st, \sqitm{x})}{\memD}{\memG}
-}
-
-\inferrule[n-syn-fail]{
-  n \hookrightarrow e \in \memG\\
-  [\Upsilon \vdash_\Phi e \nRightarrow]
-}{
-  \serr{\ssyn{n}}{\memD}{\memG}{\boxdot; \Upsilon; \Phi}
-}
-
-\inferrule[n-ana]{
-  \snorm{\st}{\memD}{\memG}{\boxdot;\Upsilon;\Phi}{\st'}{\memD'}{\memG'}\\
-  n \hookrightarrow e \in \memG'\\\\
-  \eana{\Upsilon}{\Phi}{e}{\st'}{\iota}\\
-  (x~\text{fresh})\\
-  \snorm{\srep{\st'}}{\memD'}{\cdot}{\boxdot;\Upsilon;\Phi}{\sqity{\tau}}{\memD''}{\cdot}
-}{
-  \snorm{\sana{n}{\st}}{\memD}{\memG}{\boxdot; \Upsilon; \Phi}{\sqitm{x}}{\memD''}{\memG'\otimes n \hookrightarrow e : \st' \leadsto \iota/x : \tau}
-}
-
-\inferrule[n-ana-seen-ok]{
-  \snormX{\st}{\st'}\\\\
-  n \hookrightarrow e : \st' \leadsto \iota/x : \tau \in \memG'
-}{
-  \snorm{\sana{n}{\st}}{\memD}{\memG}{\aCtx}{\sqitm{x}}{\memD'}{\memG'}
-}
-
-\inferrule[n-ana-seen-err]{
-  \snormX{\st}{\st'}\\\\
-  n \hookrightarrow e : \st'' \leadsto \iota/x : \tau \in \memG'\\
-  \st' \neq \st''
-}{
-  \serr{\sana{n}{\st}}{\memD}{\memG}{\aCtx}
-}
-
-\inferrule[n-ana-fail]{
-  \snorm{\st}{\memD}{\memG}{\boxdot;\Upsilon;\Phi}{\st'}{\memD'}{\memG'}\\
-  n \hookrightarrow e \in \memG'\\\\
-  [\Upsilon \vdash_\Phi e \nLeftarrow \st']
-}{
-  \serr{\sana{n}{\st}}{\memD}{\memG}{\boxdot; \Upsilon; \Phi}
-}
-
-\inferrule[n-raise]{ }{\serrX{\sraise{\kappa}}}
-\end{mathpar}
-\caption{Static Normalization}
-\label{dynamics-SL}
-\end{figure*}
-%\acks
-%The author is grateful to Jonathan Aldrich, Robert Bocchino and anonymous referees for their useful suggestions.
-This work was funded by the DOE Computational Science Graduate Fellowship under grant number DE-FG02-97ER25308.
-%Acknowledgments, if needed.
-% \appendix
-% \section
-% \newpage
-% \begin{figure}
-% \small
-% {\color{gray} (omitted rules generically recursive)} \hfill\fbox{$\kok{\kDelta\minput}{\kTheta\minput}{\kappa\minput}$}
-% \begin{mathpar}
-% \inferrule[kf-a]{\kalpha \in \kDelta}{\kok{\kDelta}{\kTheta}{\kalpha}}
-% ~~~~
-% \inferrule[kf-univ]{
-%   \kok{\kDelta, \kalpha}{\kTheta}{\kappa}
-% }{
-%   \kok{\kDelta}{\kTheta}{\kforall{\kalpha}{\kappa}}
-% }
-% ~~~~
-% \inferrule[kf-k]{\kvar \in \kTheta}{\kok{\kDelta}{\kTheta}{\kvar}}
-% ~~~~
-% \inferrule[kf-ind]{
-%   \kok{\kDelta}{\kTheta, \kvar}{\kappa}\\
-%   \kpos{\kvar}{\kappa}
-% }{
-%   \kok{\kDelta}{\kTheta}{\kmu{\kvar}{\kappa}}
-% }
-% \end{mathpar}
-% {\color{gray} (if $\kappa \neq \karrow{\kappa_1}{\kappa_2}$, positivity check is generically recursive)}\hfill\fbox{$\kpos{\kvar\minput}{\kappa\minput}$}
-% \begin{mathpar}
-% \inferrule[k-arrow-pos]{\kappa_1 \neq \kvar\\\kpos{\kvar}{\kappa_1}\\\kpos{\kvar}{\kappa_2}}{\kpos{\kvar}{\karrow{\kappa_1}{\kappa_2}}}
-
-% \inferrule[k-kvar-pos]{ }{\kpos{\kvar}{\kvar}}
-% \end{mathpar}
-% ~\hfill\fbox{$\kctxok{\kDelta}{\kGamma}$}\\
-% \vspace{-20px}\begin{mathpar}
-% \inferrule[kctx-emp]{ }{\kctxok{\kDelta}{\emptyset}}
-% ~~~~~~~~~~
-% \inferrule[kctx-ext]{
-%   \kctxok{\kDelta}{\kGamma}\\
-%   \kok{\kDelta}{\emptyset}{\kappa}
-% }{\kctxok{\kDelta}{\kGamma, \sx : \kappa}}
-% \end{mathpar}
-% \caption{Kinds, kinding contexts and equality kinds.}
-% \label{statics-SL2}
-% \end{figure}
+\appendix
+\section{Appendix}
+
 \end{document}

### 2014-10-02 sanitation-psp14/sanitation-psp14-short.tex
Updated intro per reviewer comments

--- 2014-08-21 sanitation-psp14/sanitation-psp14-short.tex
@@
   \and
-  \affaddr{Carnegie Mellon University}\\
-  \affaddr{Pittsburgh, PA}\\
+  \affaddr{Carnegie Mellon University, Pittsburgh, PA}\\
+  %\affaddr{Pittsburgh, PA}\\
   \email{\{nathanfu, comar, aldrich\}@cs.cmu.edu}
@@
 Alternatively, it might HTML-encode special characters, which would head off both SQL injection attacks and cross-site scripting attacks.
-Guaranteeing that user input has already been sanitized  before it is used to construct a command is challenging, because  sanitization is often performed in a different part of a program than where commands are ultimately constructed.% Note, for example, that \verb|sanitize| is not idempotent, so it should only be called once.
+Guaranteeing that user input has already been sanitized  before it is used to construct a command is challenging, especially because  sanitization and command construction may occur in separately developed and maintained components.
+%is often performed in a different component of a program than where commands are ultimately constructed.% Note, for example, that \verb|sanitize| is not idempotent, so it should only be called once.
 
@@
 These guarantees require run-time checks only when going from less precise to more precise types (e.g.
-at the edges of the system).
+along the interfaces with external systems or components that are not under the control of this type system).
 %implementing and statically checking input sanitation techniques.
@@
 \renewcommand{\grammarlabel}[2]{#1\hfill#2}
-\begin{figure}[h]
+\begin{figure}[t]
 \small
@@
 \end{figure}
-
-\begin{figure}[h]
+\begin{figure}[t]
 \small
   \begin{grammar}
-
-<$\sigma$> ::=  $\sigma \rightarrow \sigma$     \hfill  source types          \alt
-$\stringin{r}$
+<$\sigma$> ::=  $\sigma \rightarrow \sigma$ | $\stringin{r}$    \hfill  source types
 
 <$e$> ::= 
-      $x$ \hfill source terms \alt 
-      $\lambda x . e$ \alt
-      $e(e)$ \alt
-      $\strin{s}$ \hfill $s \in \Sigma^{*}$ \alt
-      $\rsconcat{e}{e}$ | $\strcase{e}{e}{x,y.e}$ \alt
-      $\sisubst{r}{e}{e}$ \alt
-      $\rcoerce{r}{e}$ | $\rcheck{r}{e}{x.e}{e}$
-
-      <$v$> ::= $\lambda x . e$ | $\strin{e}$ | $s$ \hfill source values 
+      $x$ | $\lambda x . e$ | $e(e) $\hfill source terms \alt 
+      $\strin{s}$ | $\rsconcat{e}{e}$ | $\strcase{e}{e}{x,y.e}$ \hfill $s \in \Sigma^{*}$ \alt
+      $\sisubst{r}{e}{e}$ | $\rcoerce{r}{e}$ | $\rcheck{r}{e}{x.e}{e}$
+
+<$v$> ::= $\lambda x . e$ | $\strin{s}$ \hfill source values 
 \caption{Syntax of $\lambda_{RS}$.}
@@
 
-\begin{figure}[h]
+\begin{figure}[t]
 \small
@@
 
-<$\tau$> ::= $\tau \rightarrow \tau$ \hfill target types \alt
-$\str$ \alt $\regex$
-
-<$\iota$> ::= $x$  \hfill target terms \alt
-             $\lambda x . \iota$ \alt
-                    $\iota(\iota)$ \alt
-$\tstr{s}$ \alt
-  $\rx{r}$ \alt
-  $\tconcat{\iota}{\iota}$ | $\pstrcase{\iota}{\iota}{x,y.\iota}$ \alt
-  $\preplace{\iota}{\iota}{\iota}$ | $\tcheck{\iota}{\iota}{\iota}{\iota}$
+<$\tau$> ::= $\tau \rightarrow \tau$ | $\str$ | $\regex$ \hfill target types
+
+<$\iota$> ::= $x$ | $\lambda x . \iota$ | $\iota(\iota)$ \hfill target terms \alt
+$\tstr{s}$ | $\tconcat{\iota}{\iota}$ | $\pstrcase{\iota}{\iota}{x,y.\iota}$ \alt
+  $\rx{r}$ | $\preplace{\iota}{\iota}{\iota}$ | $\tcheck{\iota}{\iota}{\iota}{\iota}$
 
@@
 Unlike other approaches, our solution complements existing, familiar solutions while providing a strong guarantee that traditional library and framework-based approaches are implemented and utilized correctly.
+
+\section{Acknowledgements}
+\todo{funding sources}
 

### 2014-10-03 sanitation-psp14/sanitation-psp14-short.tex
Minor changes.
Adeed a line to sec. 3 per review 2.

@@
 \item $\lhead{a r'} = a$ where $a \in \Sigma$
+\item $\lhead{r_1r_2} = \lhead{r_1}$
 \item $\lhead{\epsilon} = \epsilon$
@@
 
-The chief benefit of $\lambdas$ is its safety theorem, which states that any value of
+The chief benefit of $\lambdas$ is its security theorem, which states that any value of
 a regular expression type is recognized by the regular expression corresponding to its type.
-Our main technical result, stated later in this section, establishes that this property
+This is beneficial because safe subsets of strings may be characterized by regular expressions.
+Our main technical result, stated later in this section, establishes that the security property
 is preserved under translation into $\lambdap$.

### 2014-10-03 sanitation-psp14/sanitation-psp14-short.tex
Cleaned up punctuation around thms and removed the wyvern cite, which we can hopefully add back in.

@@
 
-\begin{prop}[Closure.] \label{thm:total}
+\begin{prop}[Closure] \label{thm:total}
   If $\lang{r}, \lang{r_1}$ and $\lang{r_2}$ are regular expressions, then $\lang{\lreplace{r}{r_1}{r_2}}$ is also a regular language.
@@
 
-\begin{prop}[Substitution Correspondence.] \label{thm:substcorrespondence}
+\begin{prop}[Substitution Correspondence] \label{thm:substcorrespondence}
   If $s_1 \in \lang{r_1}$ and $s_2 \in \lang{r_2}$ then
@@
 
-\begin{thm}[Canonical Forms for String Fragment of $\lambdas$.]\label{thm:cfs}
+\begin{thm}[Canonical Forms for String Fragment of $\lambdas$]\label{thm:cfs}
   If $\sctx \vdash e : \stringin{r}$ 
@@
 
-\begin{thm}[Correctness of Input Sanitation for $\lambdas$.]\label{thm:scorrect}
+\begin{thm}[Correctness of Input Sanitation for $\lambdas$]\label{thm:scorrect}
   If  $\sctx \vdash e : \stringin{r}$ and $e \sreduces \rstr{s}$ then $s \in \lang{r}$.
@@
 
-\begin{lem}
+\begin{lem}[Well-formedness of regular expressions]
 If $\tctx \vdash \iota : \regex$ then $\iota \treduces \rx{r}$ such that $r$ is a well-formed regular expression. 
@@
 
-\begin{thm} 
+\begin{thm}[Safety for $\lambdap$]
   Let $\iota$ be a term in the target language. If $\tctx \vdash \iota : \tau$ 
@@
 
-\begin{thm}[Correctness of Input Sanitation for Translated Terms.]\label{thm:main}
+\begin{thm}[Correctness of Input Sanitation for Translated Terms]\label{thm:main}
   If $\trden{e} = \iota$ and $\sctx \vdash e : \stringin{r}$ then $\iota \sreduces \tstr{s}$
@@
 
-The Wyvern programming language provides a general framework for composing language extensions \cite{tsl}\cite{wyvern}.
+The Wyvern programming language provides a general framework for composing language extensions \cite{tsl}.%\cite{wyvern}.
 Our work identifies one particular extension, and is therefore complementary to Wyvern and related

### 2014-10-06 sanitation-psp14/sanitation-psp14-short.tex
typo
Updates to Sec2 of PSP14

@@
 -- a sequence of characters other than quotation marks and backslashes; these can only appear escaped.
-This concrete syntax for regular expression patterns can be understood to desugar, in a standard way, to the syntax for regular expressions shown in Figure \ref{fig:regex}, where $r \cdot r$ is sequencing and $r + r$ is  disjunction.
+This syntax for regular expression patterns can be understood to desugar, in a standard way, to the syntax for regular expressions shown in Figure \ref{fig:regex}, where $r \cdot r$ is sequencing and $r + r$ is  disjunction.
 We will work with this ``core'' for simplicity.
@@
 \item Section \ref{sec:p} describes a simple target language, $\lambdap$, with a minimal regular expression library. In Section \ref{atlang}, we will take Python to be such a language.
-\item Section \ref{sec:tr} describes the translation from $\lambdas$ to $\lambdap$ and ensures the correctness result for \ref{sec:rs} is preserved under translation. 
+\item Section \ref{sec:tr} describes the translation from $\lambdas$ to $\lambdap$ and ensures the correctness result for \ref{sec:rs} is preserved under this translation. 
 \end{itemize}
@@
 In this section, we define a minimal typed lambda calculus with regular string types called $\lambda_{RS}$.
-The syntax of $\lambda_{RS}$ is specified in Figure \ref{fig:glambdas}, its static semantics in Figure \ref{fig:slambdas} and its evaluation semantics in Figure \ref{fig:dlambdas}.\footnote{For convenience, a single sheet containing Figures \ref{fig:glambdas}-\ref{fig:tr} is available at http://nfulton.org/strings/printout.pdf.}
+The syntax of $\lambda_{RS}$ is specified in Figure \ref{fig:glambdas}, its static semantics in Figure \ref{fig:slambdas} and its evaluation semantics in Figure \ref{fig:dlambdas}.\footnote{For convenience, a single sheet containing Figures \ref{fig:glambdas}-\ref{fig:tr} is available in the accompanying technical report \todo{TR}.}
 
@@
 $s \in \Sigma^\star$, defined in the usual way).
-The rule \textsc{S-T-Stringin-I} statically guarantees that $s \in \lang{r}$.
+The rule \textsc{S-T-Stringin-I} requires that $s \in \lang{r}$.
  %The remaining operations on terms of type $\stringin{r}$ preserve this property. %The remaining operators operate on regular strings.
@@
 The \textsc{S-T-Concat} rule is the simplest example of our approach.
-The rule is sound because the result of concatenating two strings will always be in the sequential composition two regular expressions matching the respective strings.
+The rule is sound because the result of concatenating two strings, the first in $\lang{r_1}$ and the second in $\lang{r_2}$, will always be in the language $\lang{r_1\cdot r_2}$.
 The rule therefore relates string concatenation to sequential composition of regular expressions.
 
-Whereas concatenation allows the construction of large strings from smaller strings, \textsc{S-T-Case} allows the decomposition, or elimination, of large strings into smaller strings.
-Intuitively, this rule ``peels off" the first character, and then performs a specified operation on the first character and the remaining string.
-
-In terms of strings, \textsc{S-T-Case} is essentially the ubiquitious substring operation.
-
-The definition of \textsc{S-T-Case} is subtle because even simple string operations sometimes require complicated operations on regular expressions.
-Peeling off the first character of a string is deterministic -- the first character is always one, concrete value.
-However, a regular expression describing the first character of any matching string might recognize multiple concrete values.
-For instance, the first character of ``CMU" is always ``C"; however, if the expression associated with this string is CMU$+$KIT, then the expression describing the first character is C$+$K.
-Therefore, an operation on languages corresponding to case analysis on strings must consider one-character prefixes (a head) and their corresponding suffixes (a tail).
-
-Fortunately, regular expression derivatives \cite{bowzer} conveniently capture this intuition.
+Whereas concatenation allows the construction of large strings from smaller strings,  $\strcase{e}{e_0}{x,y.e_1}$  allows the decomposition, or elimination, of large strings into smaller strings.
+Intuitively, this operation branches based on whether a string is empty or not, exactly analagous to case analysis on lists in a functional language.
+The branch for a non-empty string ``peels off" the first character, binding it and the remainder of the string to specified variables.
+The evaluation rules \textsc{S-E-Case-$\epsilon$} and \textsc{S-E-Case-Concat} express this semantics.
+This construct can be used to implement a standard string indexing operation as a function, given a suitable definition of natural numbers (omitted).
+
+The regular expressions describing the first character and remaining characters of a matching string in the language $\lang{r}$ must be determined to correctly specify the static semantics.
+For instance, the first character of a string in the language $\lang{(A\cdot B\cdot C)+(D\cdot E\cdot F)}$ is in the language $\lang{A+D}$ and the remaining characters are in the language $\lang{(B\cdot C)+(E \cdot F)}$.
+
 The regular expression recognizing any one-character prefix of the strings in $\lang{r}$ is easily defined.
-The one-character prefix definition combined with derivatives gives a concise definition of the
-remaining string.
 
@@
 
-Given this definition of $\lhead{r}$, regular expression derivatives provide a useful tool for defining $\ltail{r}$.
+Given this definition of $\lhead{r}$, regular expression derivatives \cite{bowzer} provide a useful tool for defining $\ltail{r}$.
 \begin{defn}[Brzozowski's Derivative]\label{def:derivative}
@@
 
-The \textsc{S-T-Case-Concat} rule, which is defined in terms of these operations, relates the result of ``peeling off" the first character of a string to regular expression derivatives.
+The \textsc{S-T-Case} rule, which is defined in terms of these operations, thus relates the result of ``peeling off" the first character of a string to regular expression derivatives.
 
@@
 
-The $\lambdas$ language supports two forms of coercion. Safe coercion only allows coercion between 
-strings types which are guaranteed to be safe.
-This form of coercion is useful because the replacement operation (introduced below in Section~\ref{sec:replace}) is often more conservative than absolutely necessary.
-Conversely, unsafe coercion -- which we refer to as a checked coercion -- 
-allows for potential unsafe type casts. Our semantics for check ensures that only safe values are used. Whenever a value is determined to be unsafe at runtime, a default value $e_2$ is used instead.
-
-Summarily, 
- \textsc{S-T-SafeCoerce} allows only safe coercions between string types by
-    expoiting the decidability of language inclusion, while
-\textsc{S-T-Check} allows casts between strings that cannot be guaranteed at compile time, but inserts a runtime check.
-
-An alternative to coercion is subtyping for constrained string types, which applies coercion implicitly. 
+The $\lambdas$ language supports two forms of coercion. Safe coercions, written $\rcoerce{r}{e}$, allow passage to a ``smaller'' regular language. Such coercions will always succeed. 
+% This form of coercion is useful because the replacement operation (introduced below in Section~\ref{sec:replace}) is often more conservative than absolutely necessary.
+Conversely, checked coercions, written $\rcheck{r}{e_0}{x.e_1}{e_2}$, 
+allow for passage to regular languages that are not necessarily smaller. Checked coercions branch based on whether the coercion succeeded or not.
+
+The rule 
+ \textsc{S-T-SafeCoerce} checks for language inclusion, which we write $\lang{r_1} \subseteq \lang{r_2}$.
+Language inclusion is a decidable relation.
+As a result, the rule \textsc{S-E-SafeCoerce} does not need to perform any checks.
+The rule
+\textsc{S-T-Check} does not perform any static checks on the two languages involved, only checking that the two branches have the same type.
+The checks are performed dynamically, shown in the rules \textsc{S-E-Check-Ok} and \textsc{S-E-Check-NotOk}.
+
+In our calculus, both forms of coercion are explicit.
+For safe coercions, it is often useful for the coercion to be performed implicitly.
+This can be seen as a form of  subtyping for regular string types \todo{cite?}.
 In practice, subtyping is crucial to the usability of our system due to the nature of the replacement operator.
-For simplicity, we do not include this relation; however, subtyping for constrained strings is present in \cite{fulton12,fulton13} and that treatment extends to this system.
-The implementation discussed in Section \ref{atlang} provides subtyping between constrained string types.
+For simplicity, we do not include this relation; however, subtyping for regular strings is present in previous treatments of our work \cite{fulton12,fulton13}.
+The implementation discussed in Section \ref{atlang} also provides subtyping between regular string types.
 
 \subsubsection{Replacement}\label{sec:replace}
-
-The premier operation for manipulating strings in $\lambdas$ is string substitution.
-On strings, the string substitution operator is familiar, and roughly analogous to \lstinline{str_replace} in PHP and \lstinline{String.replace} in Java.
-The definition of \textsc{S-T-Replace} requires a definition of replacement for regular expressions (or languages) corresponding to string replacement.
-In this section, we define this operation.
-
-Both string and language replacement are defined extra-linguistically.
-The system $\lambdas$ is defined in terms of these functions.
-The function $\lsubst{r}{s_1}{s_2}$ replaces every substring of $s_1$ matching $r$ with $s_2$.
-In an analogous manner, the function $\lreplace{r}{r_1}{r_2}$ replaces every sublanguage of $r_1$ matching $r$ with $r_2$.
-If the intuition for this operation is not clear, it may be helpful to think in terms of replacing sub-automata.
-
-Throughout this section, we fix an alphabet $\Sigma$ over which strings $s$ and
-regular expressions $r$ are defined. We use $\lang{r}$ to refer to the
-language recognized by the regular expression $r$.
-
-\begin{lem}[Properties of Regular Languages and Expressions.] \label{thm:regexprops}
-The following are properties of regular expressions which are necessary for our proofs:
-If $s_1 \in \lang{r_1}$ and $s_2 \in \lang{r_2}$ then $s_1s_2 \in \lang{r_1r_2}$.
-For all strings $s$ and regular expressions $r$, either $s \in \lang{r}$ or $s \not \in \lang{r}$.
-Regular languages are closed under reversal.
-\end{lem}
-
-If any of these properties are unfamiliar, the reader may refer to a standard text on the subject \cite{cinderella}.
+The premier operation for working with regular strings in $\lambdas$ is  replacement, written $\sisubst{r}{e_1}{e_2}$. 
+It behaves analogously to \lstinline{str_replace} in PHP and \lstinline{String.replace} in Java, differing in that the replacement pattern $r$ must be statically given.
+The evaluation rule \textsc{S-E-replace} is defined via the relation $\lsubst{r}{s_1}{s_2}=s$.
+For simplicity, we do not inductively define this standard relation.
 
 \begin{defn}[$\tt{subst}$]
-The relation $\lsubst{r}{s_1}{s_2} = s$ produces a string $s$ in which all substrings of $s_1$ matching $r$ are replaced with $s_2$.
+The relation $\lsubst{r}{s_1}{s_2} = s$ produces a string $s$ in which all substrings of $s_1$ in $\lang{r}$ are replaced with $s_2$.
 \end{defn}
 
-A proper definition of ${\sf lreplace}$ would give an rewrite system with correctness and termination proofs, which is beyond the scope of this paper
-on security-motivated type system extensions. Instead, we provide an abstract definition of the relation and state necessary propositions.
+The typing rule \textsc{S-T-Replace} requires computing the regular language of the string resulting from this operation given the languages of $s_1$ and $s_2$, written as the relation $\lreplace{r}{r_1}{r_2}={r'}$.
+Given an automata-oriented interpretation of regular languages, it may be helpful to think in terms of replacing sub-automata.
+
+A complete definition of ${\sf lreplace}$ would also give an rewrite system with correctness and termination proofs, which is beyond the scope of this paper.
+Instead, we provide an abstract definition of the relation and state necessary properties.
 
@@
 \begin{prop}[Closure] \label{thm:total}
-  If $\lang{r}, \lang{r_1}$ and $\lang{r_2}$ are regular expressions, then $\lang{\lreplace{r}{r_1}{r_2}}$ is also a regular language.
+  If $\lang{r}, \lang{r_1}$ and $\lang{r_2}$ are regular languages, then $\lang{\lreplace{r}{r_1}{r_2}}$ is also a regular language.
 \end{prop}
@@
 
-  $\lsubst{r}{s_1}{s_2} \in \lang{\lreplace{r}{s_1}{s_2}}$.
+  $\lsubst{r}{s_1}{s_2} \in \lang{\lreplace{r}{r_1}{r_2}}$.
 \end{prop}
@@
 of regular expressions.
+
+\begin{lem}[Properties of Regular Languages and Expressions.] \label{thm:regexprops}
+The following are properties of regular expressions which are necessary for our proofs:
+If $s_1 \in \lang{r_1}$ and $s_2 \in \lang{r_2}$ then $s_1s_2 \in \lang{r_1r_2}$.
+For all strings $s$ and regular expressions $r$, either $s \in \lang{r}$ or $s \not \in \lang{r}$.
+Regular languages are closed under reversal.
+\end{lem}
+
+If any of these properties are unfamiliar, the reader may refer to a standard text on the subject \cite{cinderella}.
 

### 2014-10-07 sanitation-psp14/sanitation-psp14-short.tex
Updated Sec 2.1.4

@@
       $\strin{s}$ | $\rsconcat{e}{e}$ | $\strcase{e}{e}{x,y.e}$ \hfill $s \in \Sigma^{*}$ \alt
-      $\sisubst{r}{e}{e}$ | $\rcoerce{r}{e}$ | $\rcheck{r}{e}{x.e}{e}$
+      $\rcoerce{r}{e}$ | $\rcheck{r}{e}{x.e}{e}$ | $\sisubst{r}{e}{e}$
 
@@
 
+\inferrule[S-T-SafeCoerce]
+{\sctx \vdash e : \stringin{r'} \\ \lang{r'} \subseteq \lang{r}}
+{\sctx \vdash \coerce{r}{e} : \stringin{r}}
+
+\inferrule[S-T-Check]
+{\sctx \vdash e_0 : \stringin{r_0} \\ \sctx, x:\stringin{r} \vdash e_1 : \sigma \\ \sctx \vdash e_2 : \sigma}
+{\sctx \vdash \rcheck{r}{e_0}{x.e_1}{e_2} : \sigma}
+
 \inferrule[S-T-Replace]
@@
 {\sctx \vdash \sisubst{r}{e_1}{e_2} : \stringin{r'}}
-
-\inferrule[S-T-SafeCoerce]
-{\sctx \vdash e : \stringin{r'} \\ \lang{r'} \subseteq \lang{r}}
-{\sctx \vdash \coerce{r}{e} : \stringin{r}}
-
-\inferrule[S-T-Check]
-{\sctx \vdash e_0 : \stringin{r_0} \\ \sctx, x:\stringin{r} \vdash e_1 : \sigma \\ \sctx \vdash e_2 : \sigma}
-{\sctx \vdash \rcheck{r}{e_0}{x.e_1}{e_2} : \sigma}
 \end{mathpar}
@@
 
-\inferrule[S-E-Replace]
-{e_1 \sreduces \strin{s_1} \\ e_2 \sreduces \strin{s_2} \\ \lsubst{r}{s_1}{s_2} = s} 
-{\sisubst{r}{e_1}{e_2} \sreduces \sistr{s}}
-
 \inferrule[S-E-SafeCoerce]
@@
 }
+
+\inferrule[S-E-Replace]
+{e_1 \sreduces \strin{s_1} \\ e_2 \sreduces \strin{s_2} \\ \lsubst{r}{s_1}{s_2} = s} 
+{\sisubst{r}{e_1}{e_2} \sreduces \sistr{s}}
 \end{mathpar}
@@
 \begin{prop}[Substitution Correspondence] \label{thm:substcorrespondence}
-  If $s_1 \in \lang{r_1}$ and $s_2 \in \lang{r_2}$ then
-
-  $\lsubst{r}{s_1}{s_2} \in \lang{\lreplace{r}{r_1}{r_2}}$.
+  If $s_1 \in \lang{r_1}$ and $s_2 \in \lang{r_2}$ then $\lsubst{r}{s_1}{s_2} \in \lang{\lreplace{r}{r_1}{r_2}}$.
 \end{prop}
 \begin{proof}
-The proposition follows from the definitions of subst and lreplace; note that language substitutions over-approximate string substitutions.
+The proposition follows from the definitions of $\textsf{subst}$ and $\textsf{lreplace}$; note that language substitutions over-approximate string substitutions.\todo{elaborate on this?}
 \end{proof}
@@
 In this section, we establish type soundness for $\lambdas$.
-The theorem relies upon the lemmas and propositions established above.
-We also rely on additional lemmas which establish the preservation of well-formedness
-of regular expressions.
-
-\begin{lem}[Properties of Regular Languages and Expressions.] \label{thm:regexprops}
-The following are properties of regular expressions which are necessary for our proofs:
-If $s_1 \in \lang{r_1}$ and $s_2 \in \lang{r_2}$ then $s_1s_2 \in \lang{r_1r_2}$.
-For all strings $s$ and regular expressions $r$, either $s \in \lang{r}$ or $s \not \in \lang{r}$.
-Regular languages are closed under reversal.
+The theorem relies upon the definitions and propositions given above and some basic properties of regular languages.
+
+\begin{lem}[Properties of Regular Languages.] \label{thm:regexprops}
+~
+\begin{enumerate}
+\item If $s_1 \in \lang{r_1}$ and $s_2 \in \lang{r_2}$ then $s_1s_2 \in \lang{r_1r_2}$.
+\item For all strings $s$ and regular expressions $r$, either $s \in \lang{r}$ or $s \not \in \lang{r}$.
+\item Regular languages are closed under reversal.
+\end{enumerate}
 \end{lem}
@@
 
+We also need to establish a well-formedness lemma, which simply expresses the totality of the operations over regular expressions used as premises in the rules in Figure \ref{fig:slambdas}. 
 \begin{lem}
@@
 \begin{proof}
-  The only non-trivial case is S-T-Replace, which follows from lemma \ref{thm:total}. 
+  The only non-trivial case is S-T-Replace, which follows from Proposition \ref{thm:total}. \todo{what about lhead/ltail?}
 \end{proof}
@@
 by our theorems about regular languages.
-We avoid the most significant issues inherent to big-step semantics by avoiding non-termination.
-The simply typed lambda calculus terminates, and our conservative, compositional extension clearly
-terminates modulo termination of ${\sf subst}$ and ${\sf lreplace}$ (but even if these were non-terminating, this still does not pose a problem for the big-step semantics itself).
-However, a more careful treatment or treatment for a language with non-termination might proceed by either coinduction \cite{coinductionoo, coinduction} or step-indexing \cite{stepindexing}.
-
-\begin{thm}[Canonical Forms for String Fragment of $\lambdas$]\label{thm:cfs}
-  If $\sctx \vdash e : \stringin{r}$ 
-  then $e \sreduces \rstr{s}$ 
-\end{thm}
+
+We avoid the most significant issues inherent to proofs related to big-step semantics by avoiding non-termination in our specification.
+The simply typed lambda calculus terminates, and our conservative extension clearly
+terminates because new binding structures are not introduced and termination (i.e. decidability) of ${\sf subst}$ and ${\sf lreplace}$ is known. Even in a non-terminating calculus, type safety holds.
+However, a more careful treatment or treatment for a language with non-termination would need to proceed by either a coinductive argument \cite{coinductionoo} or  with a small-step semantics.
 
 \begin{thm}[Type Safety.] \label{thm:scorrect}
-  If $\sctx \vdash e : \sigma$ 
-  then $e \sreduces v$ and $\sctx \vdash v : \sigma$.
+  If $\emptyset \vdash e : \sigma$ 
+  then $e \sreduces v$ and $\emptyset \vdash v : \sigma$.
 \end{thm}
@@
 By induction on the typing relation.
-The S-T-Concat case requires Lemma \ref{thm:regexprops} 
-and the S-T-Replace case appeals to Lemma \ref{thm:substcorrespondence}.
+The \textsc{S-T-Concat} case requires Lemma \ref{thm:regexprops}.1, the \textsc{S-T-Check} case requires Lemma \ref{thm:regexprops}.2 
+and the S-T-Replace case appeals to Proposition \ref{thm:substcorrespondence}.
 \end{proof}
 
+We can also define a canonical forms lemma for regular strings.
+
+\begin{lemma}[Canonical Forms for String Fragment of $\lambdas$]\label{thm:cfs}
+  If $\emptyset \vdash e : \stringin{r}$ and $e \sreduces v$ then $v=\rstr{s}$. 
+\end{lemma}
+\begin{proof}
+The proof is again by induction on the typing relation, following straightforwardly.
+\end{proof}
+
 \subsubsection{The Security Theorem}\label{sec:securitythm}
@@
 The chief benefit of $\lambdas$ is its security theorem, which states that any value of
-a regular expression type is recognized by the regular expression corresponding to its type.
-This is beneficial because safe subsets of strings may be characterized by regular expressions.
-Our main technical result, stated later in this section, establishes that the security property
-is preserved under translation into $\lambdap$.
+a regular string type is recognized by the regular language corresponding to the regular expression the type is indexed by.
+This is beneficial because this ensure that membership in a regular language known to be secure becomes manifest in the type of the string, rather than being a property that must be established extralinguistically.
 
 \begin{thm}[Correctness of Input Sanitation for $\lambdas$]\label{thm:scorrect}
-  If  $\sctx \vdash e : \stringin{r}$ and $e \sreduces \rstr{s}$ then $s \in \lang{r}$.
+  If  $\emptyset \vdash e : \stringin{r}$ and $e \sreduces \rstr{s}$ then $s \in \lang{r}$.
 \end{thm}
 \begin{proof}
-  The theorem follows directly from type safety, canonical forms for $\lambdas$, and inversion of the typing relation for $\lambdas$.
+  The theorem follows directly from type safety, canonical forms for $\lambdas$, and inversion of \textsc{S-T-Stringin-I}, the typing rule for terms of the form $\rstr{s}$.
 \end{proof}
@@
 \end{figure}
-
-The system $\lambdap$ is a straight-forward extension of a simply typed lambda calculus
-with a $\str$ type and a regular expression type, as well as 
-some operations -- such as concatenation and replacement -- found in many the standard libraries of many programming languages.
-The operations of $\lambdap $ correspond precisely to the operations of $\lambdas$ in a way made precise by the translation rules described in the next section.
-Unlike $\lambdas$, $\lambdap$ does not statically track the effects of string operations.
+Our next major technical result, stated in Sec. \ref{sec:tr}, establishes that the security property
+is preserved under translation into $\lambdap$, which we now define.
+
+The system $\lambdap$ is another straight-forward extension of a simply typed lambda calculus
+with a $\str$ type and a $\regex$ type, as well as 
+some operations -- such as concatenation and replacement -- found in the standard libraries of many programming languages.
+The operations of $\lambdap $ correspond to ``run-time'' versions of the operations performed statically in $\lambdas$ in a way made precise by the translation rules described in the next section.
+%Unlike $\lambdas$, $\lambdap$ does not statically track the effects of string operations.
 The language $\lambdap$ is so-called because it is reminscent of popular web programming languages,
-such as \textbf{P}ython or \textbf{P}HP.
+such as \textbf{P}ython or \textbf{P}HP. We will discuss an implementation within the former in the next section.
 

### 2014-10-07 sanitation-psp14/sanitation-psp14-short.tex
Updated through 2.2

@@
 \begin{enumerate}
-\item If $s_1 \in \lang{r_1}$ and $s_2 \in \lang{r_2}$ then $s_1s_2 \in \lang{r_1r_2}$.
+\item If $s_1 \in \lang{r_1}$ and $s_2 \in \lang{r_2}$ then $s_1s_2 \in \lang{r_1\cdot r_2}$.
 \item For all strings $s$ and regular expressions $r$, either $s \in \lang{r}$ or $s \not \in \lang{r}$.
@@
 
-\begin{lemma}[Canonical Forms for String Fragment of $\lambdas$]\label{thm:cfs}
+\begin{lem}[Canonical Forms for String Fragment of $\lambdas$]\label{thm:cfs}
   If $\emptyset \vdash e : \stringin{r}$ and $e \sreduces v$ then $v=\rstr{s}$. 
-\end{lemma}
+\end{lem}
 \begin{proof}
@@
 $\tstr{s}$ | $\tconcat{\iota}{\iota}$ | $\pstrcase{\iota}{\iota}{x,y.\iota}$ \alt
-  $\rx{r}$ | $\preplace{\iota}{\iota}{\iota}$ | $\tcheck{\iota}{\iota}{\iota}{\iota}$
-
-  <$\dot{v}$> ::= $\lambda x . \iota$ | \tstr{s} | \rx{r} \hfill target values
+  $\rx{r}$ | $\tcheck{\iota}{\iota}{\iota}{\iota}$ | $\preplace{\iota}{\iota}{\iota}$
+
+  <$\dot{v}$> ::= $\lambda x . \iota$ | $\tstr{s}$ | $\rx{r}$ \hfill target values
 
@@
 The language $\lambdap$ is so-called because it is reminscent of popular web programming languages,
-such as \textbf{P}ython or \textbf{P}HP. We will discuss an implementation within the former in the next section.
+such as \textbf{P}ython or \textbf{P}HP, albeit statically typed. We will discuss an implementation within the former in the next section.
 
@@
 and a big-step semantics is defined by the rules \textsc{P-E-} in Figure~\ref{fig:dlambdap}.
+The semantics are straightforward given the discussion of the corresponding operations in the previous section.
 
@@
 correctness of our translation.
-Again, we elide a more careful treatment by treating a special case where our language terminates but
-note the multiple approaches to proving soundness for non-terminating languages with natural semantics \cite{coinductionoo, stepindexing, coinduction}.
-
-\begin{lem}[Well-formedness of regular expressions]
-If $\tctx \vdash \iota : \regex$ then $\iota \treduces \rx{r}$ such that $r$ is a well-formed regular expression. 
+Again, we elide a more careful treatment by treating a special case where our language terminates per the discussion above.
+
+\begin{thm}[Safety for $\lambdap$] If $\emptyset \vdash \iota : \tau$ 
+  then $\iota \sreduces \dot v$ and $\tctx \vdash \dot v : \tau$.
+\end{thm}
+
+We can also define canonical forms for regular expressions and strings in the usual way:
+
+\begin{lem}[Canonical Forms for Target Language]
+~
+\begin{enumerate}
+\item If $\emptyset \vdash \iota : \regex$ then $\iota \treduces \rx{r}$ such that $r$ is a well-formed regular expression. 
+\item If $\emptyset \vdash \iota : \str$ then $\iota \treduces \tstr{s}$.
+\end{enumerate}
 \end{lem}
 
-\begin{thm}[Safety for $\lambdap$]
-  Let $\iota$ be a term in the target language. If $\tctx \vdash \iota : \tau$ 
-  then $\iota \sreduces \iota'$ and $\tctx \vdash \iota' : \tau$.
-\end{thm}
-
 \subsection{Translation from $\lambdas$ to $\lambdap$}\label{sec:tr}
 
-The translation from $\lambdas$ to $\lambdap$ is defined in figure 5.
+The translation from $\lambdas$ to $\lambdap$ is defined in Figure \ref{fig:tr}.
 The coercion cases are most interesting. If the safety of coercion in manifest in the

### 2014-10-07 sanitation-psp14/sanitation-psp14-short.tex
Finished Sec2

@@
     \makeatletter
-\usepackage{enumitem}
+%\usepackage{enumitem}
 \usepackage{enumerate}
@@
 \begin{enumerate}
-\item If $s_1 \in \lang{r_1}$ and $s_2 \in \lang{r_2}$ then $s_1s_2 \in \lang{r_1\cdot r_2}$.
-\item For all strings $s$ and regular expressions $r$, either $s \in \lang{r}$ or $s \not \in \lang{r}$.
-\item Regular languages are closed under reversal.
+\item 
+If $s_1 \in \lang{r_1}$ and $s_2 \in \lang{r_2}$ then $s_1s_2 \in \lang{r_1\cdot r_2}$.
+\item 
+For all strings $s$ and regular expressions $r$, either $s \in \lang{r}$ or $s \not \in \lang{r}$.
+\item 
+Regular languages are closed under reversal.
 \end{enumerate}
@@
 
-\begin{thm}[Type Safety.] \label{thm:scorrect}
+\begin{thm}[Type Safety.] \label{thm:typesafety}
   If $\emptyset \vdash e : \sigma$ 
@@
 \begin{enumerate}
-\item If $\emptyset \vdash \iota : \regex$ then $\iota \treduces \rx{r}$ such that $r$ is a well-formed regular expression. 
-\item If $\emptyset \vdash \iota : \str$ then $\iota \treduces \tstr{s}$.
+\item 
+If $\emptyset \vdash \iota : \regex$ then $\iota \treduces \rx{r}$ such that $r$ is a well-formed regular expression. 
+\item 
+If $\emptyset \vdash \iota : \str$ then $\iota \treduces \tstr{s}$.
 \end{enumerate}
@@
 types of the expressions, then no runtime check is inserted.
-If the safety of coercion is not manifest in the types, then a check is inserted.
-
-The translation is defined by the rules \textsc{Tr-} in Figure~\ref{fig:tr}.
-This section ultimately establishes that the security theorem for $\lambdas$ is preserved under compilation.
-
-\begin{figure}[h]
+If the safety of coercion is not manifest in the types, then a check is inserted. Note that regular strings translate to strings directly; there is no space overhead.
+
+The translation correctness theorem guarantees that the translation is type preserving and that semantics of the original code and translation coincide, following the treatment of compilation pioneered by the TIL compiler for SML \cite{tarditi+:til-OLD}.
+We apply the translation inline in judgements for concision when convenient.
+
+\begin{figure}[ht]
 \small
@@
 
+\inferrule[Tr-string]
+{ \ }
+{   \trden{\strin{s}} = \tstr{s}}
+
+\inferrule[Tr-Concat]
+{  \trden{e_1} = \iota_1 \\  \trden{e_2} = \iota_2}
+{  \trden{\rsconcat{e_1}{e_2}} = \tconcat{\iota_1}{\iota_2} }
+
 \inferrule[Tr-Case]
@@
 { \trden{ \strcase{e_1}{e_2}{x,y.e_3} } = \pstrcase{\iota_1}{\iota_2}{x,y.\iota_3} }
-
-\inferrule[Tr-string]
-{ \ }
-{   \trden{\strin{s}} = \tstr{s}}
-
-\inferrule[Tr-Concat]
-{  \trden{e_1} = \iota_1 \\  \trden{e_2} = \iota_2}
-{  \trden{\rsconcat{e_1}{e_2}} = \tconcat{\iota_1}{\iota_2} }
 
@@
 \begin{proof}
-The proof proceeds by induction on the typing relation for $e$ and an appropriate choice of $\iota$; in each case,
-the choice is the syntactic form in $\lambdap$ corresponding to the form under consideration (e.g. choose {\sf replace} when considering {\sf sreplace}).
-After the correct choice, the proof proceeds by our type safety theorems and an appeal to the induction hypothesis.
+The proof proceeds by induction on the typing relation for $e$.
+We choose an $\iota$ based on the syntactic form in $\lambdap$ corresponding to the form under consideration (e.g.
+we  choose {\sf replace} when considering {\sf sreplace}).
+The proof proceeds by our type safety theorems and an appeal to the induction hypothesis.
 \end{proof}
@@
 
-Finally, our main result establishes that correctness of $\lambdas$ is preserved under the translation into $\lambdap$.
+Finally, our main result establishes that correctness of $\lambdas$ is preserved under the translation into $\lambdap$. \renewcommand*{\proofname}{Proof}
 
 \begin{thm}[Correctness of Input Sanitation for Translated Terms]\label{thm:main}
-  If $\trden{e} = \iota$ and $\sctx \vdash e : \stringin{r}$ then $\iota \sreduces \tstr{s}$
+  If $\trden{e} = \iota$ and $\emptyset \vdash e : \stringin{r}$ then $\iota \sreduces \tstr{s}$
   for $s \in \lang{r}$.
@@
 \begin{proof}
-  By theorem \ref{thm:trcorrect}, $\iota \sreduces \tstr{s}$ implies that $e \sreduces \strin{s}$.
-  By theorem \ref{thm:scorrect}, the above property together with the assumption that $e$ is well-typed implies that $s \in \lang{r}$.
+  By Theorem \ref{thm:trcorrect}, $\iota \sreduces \tstr{s}$ implies that $e \sreduces \strin{s}$.
+  By Theorem \ref{thm:scorrect}, the above property together with the assumption that $e$ is well-typed implies that $s \in \lang{r}$.
 \end{proof}
@@
 
-\begin{figure}[h]\small\begin{lstlisting}
+\begin{figure}[ht]\small\begin{lstlisting}
 from atlib import fn, stringin

### 2014-10-07 sanitation-psp14/sanitation-psp14-short.tex
Finished initial revisions

@@
 \end{mathpar}
-\caption{Translation from source terms ($e$) to target terms ($\iota$).}
+\caption{Translation from source to target.}
 \label{fig:tr}
@@
 
+\newcommand{\lstinlinep}[1]{\lstinline[language=Python,basicstyle=\ttfamily\small]{#1}}
+
+Figure \ref{fig:atexample} demonstrates the use of two type constructors, \verb|fn| and \verb|stringin|, corresponding to the two type constructors of $\lambda_{RS}$.
+We show these as being imported from \verb|atlib|, the standard library for \verb|atlang| (it benefits from no special support from the language itself).
+
+The \verb|fn| type constructor can be used to annotate functions that should be statically checked by \verb|atlang|.\footnote{Here, we use argument annotation syntax only available in versions 3.0+ of Python.
+Alternative syntax supporting Python 2.7+ can also be used, but we omit it here.} The function \verb|sanitize| on lines 3-7, for example,  specifies one argument, \verb|s|, of type \lstinlinep{stringin[r'.*']}.
+Here, the index is a regular expression, written using Python's \emph{raw string notation} as is conventional (in this particular instance, the \verb|r| is not strictly necessary).
+The sanitize function takes an arbitrary string and returns a string without
+double quotes or left and right brackets. Note that the return type need not be specified: \verb|atlang| uses a form of \emph{local type inference} \cite{Pierce:2000:LTI:345099.345100}.
+
+In this example, we use an HTML encoding so that the same sanitization function can be used to generate both SQL commands and HTML securely.
+The sanitized string is generated by invoking the \verb|replace| operator, which has the same semantics as $\textsf{rreplace}$ in $\lambda_{RS}$.
+Unlike in the core calculus, it is invoked like a method on \verb|s|.
+The regular expression determining the substrings to be replaced is given as the first argument (as in $\lambda_{RS}$, the only restriction here is that the regular expression must be specified statically.)
+
 \begin{figure}[ht]\small\begin{lstlisting}
@@
 
-\newcommand{\lstinlinep}[1]{\lstinline[language=Python,basicstyle=\ttfamily\small]{#1}}
-
-Figure \ref{fig:atexample} demonstrates the use of two type constructors, \verb|fn| and \verb|stringin|, corresponding to the two type constructors of $\lambda_{RS}$.
-We show these as being imported from \verb|atlib|, the standard library for \verb|atlang| (it benefits from no special support from the language itself).
-
-The \verb|fn| type constructor can be used to annotate functions that should be statically checked by \verb|atlang|.\footnote{Here, we use argument annotation syntax only available in versions 3.0+ of Python.
-Alternative syntax supporting Python 2.7+ can also be used, but we omit it here.} The function \verb|sanitize| on lines 3-7, for example,  specifies one argument, \verb|s|, of type \lstinlinep{stringin[r'.*']}.
-Here, the index is a regular expression, written using Python's \emph{raw string notation} as is conventional (in this particular instance, the \verb|r| is not strictly necessary).
-The sanitize function takes an arbitrary string and returns a string without
-double quotes or left and right brackets. Note that the return type need not be specified: \verb|atlang| uses a form of \emph{local type inference} \cite{Pierce:2000:LTI:345099.345100}.
-
-In this example, we use an HTML encoding so that the same sanitization function can be used to generate both SQL commands and HTML securely.
-The sanitized string is generated by invoking the \verb|replace| operator, which has the same semantics as $\textsf{rreplace}$ in $\lambda_{RS}$.
-Unlike in the core calculus, it is invoked like a method on \verb|s|.
-The regular expression determining the substrings to be replaced is given as the first argument (as in $\lambda_{RS}$, the only restriction here is that the regular expression must be specified statically.)
-
 The functions \verb|results_query| and \verb|results_div| construct a SQL query and an HTML snippet, respectively, by regular string concatenation.
@@
 
-One somewhat subtle issue here is that the return type of \verb|sanitize| is equivalent to \lstinlinep{stringin[r'[^"<>]*']}, which is a distinct type from the argument types to \verb|results_query| and \verb|results_div|.
+One somewhat subtle issue here is that the return type of \verb|sanitize| is equivalent to \lstinlinep{stringin[r'[^\"<>]*']}, which is a distinct type from the argument types to \verb|results_query| and \verb|results_div|.
 More specifically, however, it is a ``smaller'' type, in that it could be coerced to these argument types using an operator like $\textsf{rcoerce}$ in $\lambda_{RS}$.
@@
 
-The input sanitation problem is well-understood.
+The input sanitation problem is well-studied.
 There exist a large number of techniques and technologies, proposed by both practitioners and researchers, for preventing injection-style attacks.
 In this section, we explain how our approach to the input sanitation problem differs from each of these approaches.
-More important than these differences, however, is our more general assertion that language extensibility is a promising approach toward consideration of security goals in programming language design.
-
-Unlike \emph{frameworks and libraries} provided by languages such as Haskell and Ruby, our type system provides a \emph{static} guarantee that input is always properly sanitized before use.
+Equally important, however, is our more general assertion that language extensibility is the right approach for  consideration of language-oriented security mechanisms like the one we described here.
+
+Unlike {frameworks and libraries} provided by languages such as Haskell and Ruby, our type system provides a \emph{static} guarantee that input is always properly sanitized before use.
 Doing so requires reasoning about the operations on regular languages corresponding to standard operations on strings; we are unaware of any production system which contains this form of reasoning.
-Therefore, even where frameworks and libraries provide a viable interface or wrapper around input sanitation, our approach is complementary because it ensures the correctness of the framework or library itself.
+Therefore, even where frameworks and libraries provide a viable interface or wrapper around input sanitation (e.g.
+prepared SQL statements), our approach is complementary because it can be used to ensure the correctness of that framework or library itself.
 Furthermore, our approach is more general than database abstraction layers because our mechanism is applicable to all forms of command injection (e.g.
@@
 
-A number of research languages provide static guarantees that a program is free of input sanitation vulnerabilities \cite{UrFlowOSDI10}. 
-Unlike this work, our solution to the input sanitation problem has a very low barrier to adoption; for instance, our implementation conservatively extends Python -- a popular language among web developers.
-We also believe our general approach is better-positioned for security, where continuously evolving threats might require frequent addition of new analyses; in these cases, the composability and generality of our approach is a substantial advantage.
+A number of research languages provide static guarantees that a program is free of input sanitation vulnerabilities by never using a string representation at all, but rather desugaring SQL syntax to a safe representation immediately \cite{UrFlowOSDI10}.
+The Wyvern programming language introduced a general framework for writing syntax extensions like this \cite{tsl}.
+Unlike this work, our solution to the input sanitation problem retains a string representation and thus has a very low barrier to adoption.
+Our implementation conservatively extends Python -- a popular language among web developers -- rather than requiring the adoption of a new language entirely.
+We also believe our semantic extensibility approach (as opposed to only syntactic extensibility in Wyvern) is better-positioned for security, where continuously evolving threats might require frequent addition of new semantic  analyses.
+The \verb|atlang| system is particularly well-suited to type system based analyses.
 
@@
 
-The Wyvern programming language provides a general framework for composing language extensions \cite{tsl}.%\cite{wyvern}.
-Our work identifies one particular extension, and is therefore complementary to Wyvern and related
-work on extensible programming languages.
-We are also unaware of any extensible programming languages which emphasize applications to security concerns.
-
 % HosoyaPierce2002? 
@@
 \begin{itemize}
-  \item Our system is defined within an extensible type system.
-  \item We demonstrate that regular expression types are applicable to the web security domain, whereas previous work on regular expression types focused on XML schema.
   \item Although our static replacement operation is definable in some languages with regular expression types, we are the first to expose this operation and connect the semantics of regular language replacement with the semantics of string substitution via a type safety and compilation correctness argument.
+  \item We demonstrate that regular expression types are applicable to the web security domain, whereas previous work on regular expression types focused on XML.
+  \item The underlying representation of regular strings is guaranteed to be a string, rather than a heavier implementation based on an encoding using functional datatypes.
+  \item We work within an extensible type system.
 \end{itemize}
 
-In conclusion, our contribution is a type system, implemented within an extensible type system, for checking the correctness of input sanitation algorithms.
+%In conclusion, our contribution is a type system, implemented within an extensible type system, for checking the correctness of input sanitation algorithms.
 
@@
 We believe that this type system extension serves as a useful basis for web-oriented static analysis;
-frameworks and regular expression libraries could be annotated, along with use-sites.
+frameworks and regular expression libraries could be annotated, along with use-sites. We hope to empirically evaluate the feasability of this approach in the future using \verb|atlang|. 
 %The use sites of arbitrary strings are usually constrained to a few sections of the application. Assuming
@@
 We also believe that extensible programming languages are a promising approach toward incorporating other security
-and privacy analyses into programming languages.
+ analyses into programming languages.
+Construing such analyses as type systems, specifying them rigorously and implementing them within an extensible type system appears to be a promising general technique that the community may wish to emulate.
 %languages might provide developers with a way to incorporate domain-specific security and privacy analyses into their
@@
 %
-\section{Conclusion}
-
-Composable analyses which complement existing approaches constitute a promising approach toward the integration of security concerns into programming languages.
-In this paper, we presented a system with both of these properties and defined a security-preserving transformation.
-Unlike other approaches, our solution complements existing, familiar solutions while providing a strong guarantee that traditional library and framework-based approaches are implemented and utilized correctly.
+% \section{Conclusion}
+
+% Composable analyses which complement existing approaches constitute a promising approach toward the integration of security concerns into programming languages.
+% In this paper, we presented a system with both of these properties and defined a security-preserving transformation.
+% Unlike other approaches, our solution complements existing, familiar solutions while providing a strong guarantee that traditional library and framework-based approaches are implemented and utilized correctly.
 

### 2014-10-07 sanitation-psp14/sanitation-psp14-short.tex
Updated related work slightly
Final edits.

@@
 In this section, we define a minimal typed lambda calculus with regular string types called $\lambda_{RS}$.
-The syntax of $\lambda_{RS}$ is specified in Figure \ref{fig:glambdas}, its static semantics in Figure \ref{fig:slambdas} and its evaluation semantics in Figure \ref{fig:dlambdas}.\footnote{For convenience, a single sheet containing Figures \ref{fig:glambdas}-\ref{fig:tr} is available in the accompanying technical report \todo{TR}.}
+The syntax of $\lambda_{RS}$ is specified in Figure \ref{fig:glambdas}, its static semantics in Figure \ref{fig:slambdas} and its evaluation semantics in Figure \ref{fig:dlambdas}.\footnote{For convenience, a single sheet containing Figures \ref{fig:glambdas}-\ref{fig:tr} is available in the accompanying technical report \cite{psptr}.}
 
@@
 For safe coercions, it is often useful for the coercion to be performed implicitly.
-This can be seen as a form of  subtyping for regular string types \todo{cite?}.
+This can be seen as a form of  subtyping for regular string types \cite{fulton13}.
 In practice, subtyping is crucial to the usability of our system due to the nature of the replacement operator.
@@
 \begin{proof}
-The proposition follows from the definitions of $\textsf{subst}$ and $\textsf{lreplace}$; note that language substitutions over-approximate string substitutions.\todo{elaborate on this?}
+The proposition follows from the definitions of $\textsf{subst}$ and $\textsf{lreplace}$; note that language substitutions over-approximate string substitutions.
 \end{proof}
@@
 \begin{proof}
-  The only non-trivial case is S-T-Replace, which follows from Proposition \ref{thm:total}. \todo{what about lhead/ltail?}
+  The only non-trivial cases are S-T-Replace and S-T-Concat. The first follows from Proposition \ref{thm:total}, and the other
+  follows from the fact that lhead and ltail always produce well-formed expressions.
 \end{proof}
@@
   \item Although our static replacement operation is definable in some languages with regular expression types, we are the first to expose this operation and connect the semantics of regular language replacement with the semantics of string substitution via a type safety and compilation correctness argument.
+  \item The underlying representation of regular strings is guaranteed to be a string, rather than a heavier implementation based on an encoding using functional datatypes.
   \item We demonstrate that regular expression types are applicable to the web security domain, whereas previous work on regular expression types focused on XML.
-  \item The underlying representation of regular strings is guaranteed to be a string, rather than a heavier implementation based on an encoding using functional datatypes.
   \item We work within an extensible type system.
@@
 \section{Acknowledgements}
-\todo{funding sources}
-
+This work was supported by supported by the National Security Agency
+lablet contract \#H98230-14-C-0140
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

### 2014-10-07 sanitation-psp14/sanitation-psp14-short.tex
Rearranged def 1
Final edits for PSP14

@@
 \begin{defn}[Definition of \lhead{r}]
-The function $\lhead{r}$ is defined in terms of the structure of $r$:
+The relation $\lhead{r}=r'$ is defined in terms of the structure of $r$:
 \begin{itemize}[noitemsep]
-\item $\lhead{a r'} = a$ where $a \in \Sigma$
-\item $\lhead{r_1r_2} = \lhead{r_1}$
 \item $\lhead{\epsilon} = \epsilon$
+\item $\lhead{.} = a_1 + a_2 + ... + a_n$ for all $a_i \in \Sigma$ where $|\Sigma| = n$.
+\item $\lhead{a} = a$ %where $a \in \Sigma$
+\item $\lhead{r_1\cdot r_2} = \lhead{r_1}$
 \item $\lhead{r_1 +  r_2}  = \lhead{r_1} + \lhead{r_2}$
 \item $\lhead{r*} = \lhead{r}$
-\item $\lhead{.} = a_1 + a_2 + ... + a_n$ for all $a_i \in \Sigma$ where $|\Sigma| = n$.
 \end{itemize}
@@
 \begin{defn}[Definition of \ltail{r}]
-The function \ltail{r} is defined in terms of $\lhead{r}$.
+The relation $\ltail{r} = r'$ is defined in terms of $\lhead{r}$.
 Note that $\lhead{r} = a_1 + a_2 + ... + a_i$.
@@
 The rule
-\textsc{S-T-Check} does not perform any static checks on the two languages involved, only checking that the two branches have the same type.
-The checks are performed dynamically, shown in the rules \textsc{S-E-Check-Ok} and \textsc{S-E-Check-NotOk}.
+\textsc{S-T-Check}, conversely, does not perform any static checks on the two languages involved, only checking that the two branches have the same type.
+The checks are performed dynamically, by rules \textsc{S-E-Check-Ok} and \textsc{S-E-Check-NotOk}.
 
@@
 The simply typed lambda calculus terminates, and our conservative extension clearly
-terminates because new binding structures are not introduced and termination (i.e. decidability) of ${\sf subst}$ and ${\sf lreplace}$ is known. Even in a non-terminating calculus, type safety holds.
-However, a more careful treatment or treatment for a language with non-termination would need to proceed by either a coinductive argument \cite{coinductionoo} or  with a small-step semantics.
+terminates because new binding structures are not introduced and termination (i.e.
+decidability) of ${\sf subst}$ and ${\sf lreplace}$ is known.
+Even in a non-terminating calculus, type safety would hold, but a more careful treatment or treatment for a language with non-termination would need to proceed by either a coinductive argument \cite{coinductionoo} or  with a small-step semantics.
 
@@
 \begin{thm}[Safety for $\lambdap$] If $\emptyset \vdash \iota : \tau$ 
-  then $\iota \sreduces \dot v$ and $\tctx \vdash \dot v : \tau$.
+  then $\iota \sreduces \dot v$ and $\emptyset \vdash \dot v : \tau$.
 \end{thm}
@@
 The translation correctness theorem guarantees that the translation is type preserving and that semantics of the original code and translation coincide, following the treatment of compilation pioneered by the TIL compiler for SML \cite{tarditi+:til-OLD}.
-We apply the translation inline in judgements for concision when convenient.
+Note that we apply the translation inline in judgements for concision when convenient.
 
@@
 \begin{proof}
-  By Theorem \ref{thm:trcorrect}, $\iota \sreduces \tstr{s}$ implies that $e \sreduces \strin{s}$.
-  By Theorem \ref{thm:scorrect}, the above property together with the assumption that $e$ is well-typed implies that $s \in \lang{r}$.
+  By Theorem \ref{thm:trcorrect} and the rules given, $\iota \sreduces \tstr{s}$ implies that $e \sreduces \strin{s}$.
+  Theorem \ref{thm:scorrect} together with the assumption that $e$ is well-typed implies that $s \in \lang{r}$.
 \end{proof}

### 2014-10-07 sanitation-psp14/sanitation-psp14-short.tex
Final edits for PSP14
minor edits.

--- 2014-10-07 sanitation-psp14/sanitation-psp14-short.tex
@@
 
+The accompanying technical report \cite{psptr} contains more detailed proofs, further discussion of string substitution and language replacement, and
+a single sheet containing Figures \ref{fig:glambdas}-\ref{fig:tr} for the reader's convenience.
+
 \subsection{The Language of Regular Strings}\label{sec:rs}
 In this section, we define a minimal typed lambda calculus with regular string types called $\lambda_{RS}$.
-The syntax of $\lambda_{RS}$ is specified in Figure \ref{fig:glambdas}, its static semantics in Figure \ref{fig:slambdas} and its evaluation semantics in Figure \ref{fig:dlambdas}.\footnote{For convenience, a single sheet containing Figures \ref{fig:glambdas}-\ref{fig:tr} is available in the accompanying technical report \cite{psptr}.}
+The syntax of $\lambda_{RS}$ is specified in Figure \ref{fig:glambdas}, its static semantics in Figure \ref{fig:slambdas} and its evaluation semantics in Figure \ref{fig:dlambdas}.
 
@@
 \item $\lhead{r_1 +  r_2}  = \lhead{r_1} + \lhead{r_2}$
-\item $\lhead{r*} = \lhead{r}$
+\item $\lhead{r*} = \epsilon + \lhead{r}$
 \item $\lhead{.} = a_1 + a_2 + ... + a_n$ for all $a_i \in \Sigma$ where $|\Sigma| = n$.

### 2014-10-07 sanitation-psp14/final.tex
remove -short files and added final.pdf

--- 2014-10-07 sanitation-psp14/sanitation-psp14-short.tex
@@
 \item $\lhead{r_1 +  r_2}  = \lhead{r_1} + \lhead{r_2}$
-\item $\lhead{r*} = \lhead{r}$
+\item $\lhead{r*} = \epsilon + \lhead{r}$
 \end{itemize}
@@
 This work was supported by supported by the National Security Agency
-lablet contract \#H98230-14-C-0140
+lablet contract \#H98230-14-C-0140.
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

### 2014-10-07 sanitation-psp14/final.tex
Acks

@@
 \section{Acknowledgements}
-This work was supported by supported by the National Security Agency
+We thank the anonymous reviewers for helpful comments.
+This work was supported by the National Security Agency
 lablet contract \#H98230-14-C-0140.

### 2014-10-12 atlang-modularity15/atlang-modularity15.tex
Renamed to 15

--- 2014-07-22 atlang-gpce14/atlang-gpce14.tex
@@
   morekeywords={lambda, self, assert, cls, @classmethod, with},
+  deletekeywords={min,max},
   numbers=left,
@@
 
-\title{Active Type Constructors}
-\subtitle{Unambiguously Growing a Static Type System Inside a Python}
+\title{Extensible Gradual Typing Inside a Python}
+%\subtitle{Unambiguously Growing a Static Type System Inside a Python}
 %\titlebanner{{\tt \textcolor{Red}{{\small Under Review -- distribute within CMU only.}}}}        % These are ignored unless
@@
 %abstractions, all within libraries.
-Programmers justifiably prefer abstractions available as libraries for the language they already use over those requiring a switch to a new language.
+Programmers prefer using abstractions available as libraries for the language they already use over those requiring a switch to a new language.
 But library providers cannot typically extend the type systems of these languages directly, so powerful statically typed abstractions  remain obscure.
-This suggests that a language is needed that is \emph{backwards compatible} with the libraries and tools of a widely adopted language, but that has an  \emph{internally extensible type system}, organized around a mechanism that guarantees that any combination of extensions can be imported  without  ambiguity.
-
-Our aim in this paper is to 1) introduce such a mechanism, and 2) show how it can itself be implemented as a library for an existing widely used language.
+This suggests that a language is needed that is {backwards compatible} with the libraries and tools of a widely adopted language, but that has an  {internally extensible type system}, organized around a mechanism that guarantees that any combination of extensions can be imported safely.
+
+Our aim in this paper is to 1) introduce such a mechanism; and 2) show how it can be implemented as a library for an existing widely used language, here Python.
 We describe two artifacts: @\texttt{lang}, an  extensible statically typed language implemented as a library inside Python, and @$\lambda$, a minimal core calculus.
-Both @\texttt{lang} and @$\lambda$ have a fixed syntax but  determine the semantics of each term constructor by delegating control, according to a novel bidirectional typechecking and translation protocol, to methods associated with a relevant \emph{active type constructor}, defined in a library.
-The delegate is stable even when additional extensions are imported, so we avoid ambiguities by construction.
-We demonstrate with several examples of interesting type system fragments as libraries.% By flexibly  repurposing a fixed syntax in this type-directed manner, we avoid ambiguities by construction.
+Both @\texttt{lang} and @$\lambda$ have a fixed syntax but  determine the semantics of each term by delegating control, according to a simple but powerful bidirectionally typed elaboration semantics, to methods associated with a relevant \emph{active type constructor}, defined in a library.
+This process is stable even when additional extensions are imported, so we avoid ambiguity by construction.
+We describe several examples of interesting type system fragments as libraries, including functional constructs, a type system for secure string sanitation and a foreign function interface to OpenCL.% By flexibly  repurposing a fixed syntax in this type-directed manner, we avoid ambiguities by construction.
 %In @\texttt{lang}, types and functions are defined using Python's syntax, distributed using Python's package system and can call into Python libraries, so @\texttt{lang} is uniquely backwards compatible.
@@
 
-Unfortunately, researchers and domain experts designing new abstractions can  find it difficult to soundly and naturally implement them as libraries, particularly when they require strengthening the language's static type system.
+Unfortunately, researchers and domain experts designing new abstractions sometimes find it difficult to soundly, completely and naturally implement them as libraries, particularly those that rely on specialized support from a static type system.
 In these situations, abstraction providers generally develop a new language dialect, increasingly with the aid of tools like language workbenches \cite{erdweg2013state}.
-Unfortunately, this \emph{language-oriented approach} \cite{journals/stp/Ward94} is problematic at language boundaries: working with components written in a foreign language requires programming against (and thus revealing) its internal implementation.
-This is brittle, unnatural and unsafe and can cause performance issues due to additional necessary checks and conversions at language boundaries.
-%To avoid this, it is necessary to combine the different languages and their associated tools, but this often requires significant effort and expertise (we will discuss this further in Sec.
+Unfortunately, this \emph{language-oriented approach} \cite{journals/stp/Ward94} is problematic at language boundaries: working with components written in several foreign languages often requires programming against a low-level internal implementation, rather than its semantics, via a foreign-function interface (FFI).
+This is brittle, unnatural and unsafe and can cause performance issues due to necessary checks and conversions at language boundaries.
+As more languages are used together,  more FFIs must be developed and traversed.%To avoid this, it is necessary to combine the different languages and their associated tools, but this often requires significant effort and expertise (we will discuss this further in Sec.
 \ref{related-work}).
@@
 Moreover, its tool support is more limited.
-This speaks to a widespread problem: programmers and development teams often cannot use  the typed abstractions they prefer because they are only available bundled with specialized languages they cannot  adopt \cite{Meyerovich:2012:SDR:2414721.2414724,Meyerovich:2013:EAP:2509136.2509515}.
+This appears to be a widespread problem: programmers and development teams often cannot use  the typed abstractions they prefer because they are only available bundled with specialized languages they cannot  adopt \cite{Meyerovich:2012:SDR:2414721.2414724,Meyerovich:2013:EAP:2509136.2509515}.
 %These languages become, in essence, merely vehicles for communicating ideas, rather than useful tools.
@@
 
-Internally extensible languages have promised to reduce the problems associated with distinct dialects by giving abstraction providers more direct control over a language's syntax and semantics from within libraries.
+Extensible languages have promised to reduce the problems associated with distinct dialects by giving abstraction providers more direct control over a language's syntax and semantics and supporting the composition of these language fragments within or alongside libraries.
 %In   an internally extensible language, programmers  can \emph{granularly} import the primitive constructs best suited to each component of their application or library.
@@
 Unfortunately, the mechanisms available today have several problems.
-First, they are themselves often available only within a distinct dialect of an existing language and so face the classic ``chicken-and-egg'' problem: a language like SugarJ \cite{erdweg2011sugarj} (which permits syntax extensions over a fixed semantics) must overcome the same extrinsic barriers as a language like HJ.
-Second, giving abstraction providers too much control over the language can permit the definition of ambiguous extensions, a problem that is detected ``too late'': when extensions are combined by abstraction clients, who are not necessarily  capable of understanding and fixing the ambiguity.% Too many constraints can leave it difficult to express real-world abstractions.%It is difficult to guarantee that extensions are sound in isolation and more so to guarantee that they can be {safely} combined.
+First, they are themselves often available only within a dialect of an existing language and so face the classic ``chicken-and-egg'' problem: a language like SugarJ \cite{erdweg2011sugarj} or Wyvern \cite{TSLs} (which permit only syntax extensions over a fixed semantics) must overcome the same extrinsic barriers as a language like HJ.
+Second, giving abstraction providers too much control over the language can permit the definition of conflicting  extensions, a problem that is detected ``too late'': when extensions are combined by abstraction clients, who are not necessarily  capable of understanding and fixing the conflict.% Too many constraints can leave it difficult to express real-world abstractions.%It is difficult to guarantee that extensions are sound in isolation and more so to guarantee that they can be {safely} combined.
 %This requires further research before it would be appropriate to widely rely on them (which we will discuss as we go on).
 
-\paragraph{Terminology}
-A type system is typically defined in fragments organized around (and often identified by) a \emph{type constructor} (e.g.
+% \paragraph{Terminology}
+% A type system is typically defined in fragments organized around (and often identified by) a \emph{type constructor} (e.g.
 a fragment defining binary product types, organized around the type constructor $\fvar{prod}$).
@@
 %Static terms are also called \emph{type-level terms}.
- often, the static terms are not made explicit but rather assumed to be in the ambient metatheory).%Types classify expressions according to the rules comprising the fragment's type system.
+%  often, the static terms are not made explicit but rather assumed to be in the ambient metatheory).%Types classify expressions according to the rules comprising the fragment's type system.
 A type safe language is one where evaluation of a well-typed expression cannot lead to undefined behavior or result in a value of a different type, greatly simplifying static reasoning about program behavior and eliminating the need for many run-time checks.
 
-%In a monolithic language design, decisions about which fragments are to be combined to constitute a complete  language are made centrally, so the designer can ensure that the concrete syntax is unambiguous and that other metatheoretic concerns, like guarantees of type safety, are maintained.
+% %In a monolithic language design, decisions about which fragments are to be combined to constitute a complete  language are made centrally, so the designer can ensure that the concrete syntax is unambiguous and that other metatheoretic concerns, like guarantees of type safety, are maintained.
 To decentralize control over these decisions, so that fragments can be defined as libraries and combined arbitrarily, these guarantees must be shown to hold for all possible combinations of fragments deemed individually valid.
@@
 
-  Each fragment also defines rules that contribute to the \emph{static semantics} (specifying how types are assigned to expressions at ``compile-time'') and the \emph{dynamic semantics} (specifying the behavior of well-typed expressions at ``run-time'') \cite{pfpl}.
+%   Each fragment also defines rules that contribute to the \emph{static semantics} (specifying how types are assigned to expressions at ``compile-time'') and the \emph{dynamic semantics} (specifying the behavior of well-typed expressions at ``run-time'') \cite{pfpl}.
 The dynamic semantics of full-scale languages are generally implemented (or even directly specified, as in the Harper-Stone semantics for Standard ML \cite{Harper00atype-theoretic}) by elaboration to a simpler \emph{internal language}.% In some languages, static terms themselves may have their own static and dynamic semantics (the \emph{static statics} and the \emph{static dynamics}).
 
-The most apparently difficult problems related to combining fragments to form a typed programming language arise when attempting to enable the decentralized introduction of new concrete forms or term constructors.
-Syntactic ambiguities are difficult to modularly preclude and many tools operate by exhaustive case analysis over the concrete or abstract syntax of the terms of the language they work with (e.g.
+Our focus in this paper is to enable the introduction of new static type system fragments as safely composable libraries for an existing, widely-used language.
+The most immediately difficult problems related to composing type system fragments arise when attempting to enable the decentralized introduction of new concrete syntax or corresponding term constructors to support the operations of this new type system.
+Syntactic ambiguities are difficult to modularly preclude (though some techniques exist).
+Moreover, many tools operate by exhaustive case analysis over the concrete or abstract syntax of the terms of the language they work with (e.g.
 syntax highlighters, editor modes, pretty-printers, style checkers, documentation generators).
@@
 In this paper, we aim to show that this need not be the case.
-We introduce a method that allows library providers to introduce new type system fragments by defining new type constructors in libraries while leaving the concrete syntax, the term constructors and the internal language fixed and enriching the static language so that it has its own dynamic semantics (the \emph{static dynamics}).
-The key idea is that we determine the semantics of each term not based directly on the identity of its term constructor (the usual ``syntax-directed'' approach), but rather by delegating to a static function associated with a relevant type constructor.
+We introduce a method that allows library providers to introduce new type system fragments by defining only new type constructors, in libraries, while leaving the concrete syntax, the term constructors and the target language of compilation fixed.
+The key idea is that we determine the semantics of each term not based directly on the identity of its term constructor (the usual ``syntax-directed'' approach), but rather by delegating to a statically evaluated function associated with a relevant type constructor.
 The term constructor's identity only serves to determine how the delegate is assigned.
@@
 
-As a simple example, consider the concrete term \lstinlinep{e.r} (where \lstinlinep{e} is an expression and \lstinlinep{r} is a static label) corresponding to the now generalized abstract term  $\FF{attr}[\text{``r''}](e)$.
-Our semantics delegate responsibility for this term to a static function associated with the type constructor of the type recursively synthesized by the ``target'', $e$.
-As a result, tuple projection, as described above, can be introduced without the need to consider that it might syntactically coincide with projection out of a functional record or an object with a field labeled \lstinlinep{r} (both examples we will discuss).
-Each of these abstractions can use the same form and term constructor.
+As a simple example, consider the concrete term \lstinlinep{e.r} (where \lstinlinep{e} is an expression and \lstinlinep{r} is a static label).
+A fragment defining statically typed pairs might want this to refer to projection of the right component of the pair, while a fragment defining functional record types might want it project out the field labeled \verb|r| and an object system might introduce a more complex semantics involving dynamic dispatch.
+Were we to naively ``open up'' the syntax and semantics of the language, these fragments would conflict.
+
+Instead, we assume that the syntax is fixed and includes this commonly used form, mapping it to abstract term  $\FF{attr}[\text{``r''}](e)$ (i.e.
+the term constructor $\FF{attr}$, \emph{indexed} by a static representation of the label and taking one \emph{argument}, $e$).
+Our semantics delegate responsibility for this term to a static function associated with the type constructor of the type recursively synthesized by the ``target'' of the operation, $e$.
+Tuples, records, objects and variants thereof can all be supported over a common syntax.
 As we will detail, terms for which there is no clear ``target'' from which to determine a delegate (e.g.
-number or list literals) instead either use the type they are being analyzed against or, if not available, a ``default'' delegate determined by the type constructor governing the function in which they appear.
+number or list literals) instead either use the type they are being analyzed against (consistent with how extensible parsing is done in Wyvern \cite{TSLs}) or, if not available, a ``default'' delegate determined by the type constructor governing the function in which they appear.
 Symmetric binary terms (e.g.
 addition) use a third protocol, which we will discuss.
-There is always exactly one active type constructor delegated  responsibility over a term and this delegate is stable with respect to additional imports, so semantic ambiguities cannot arise by construction.
+There is always exactly one active type constructor delegated  responsibility over a term and this delegate is stable with respect to additional extension imports, so semantic ambiguities cannot arise by construction.
 %This is a generalization of a previous mechanism that we have developed, which deals only with extending the elaboration, but not the typechecking, of the bodies of delimited literals \cite{omar-ecoop14}.
@@
 Our target audience is typed language fragment providers (researchers or domain experts) and language designers interested in extensibility mechanisms.
-We do not aim to evaluate the claim that our approach will ultimately be of use to fragment clients (i.e.
+% We do not aim to evaluate the claim that our approach will ultimately be of use to fragment clients (i.e.
 normal developers) in writing and reasoning about a variety of real-world programs (this first requires adoption by our target audience).
@@
 \begin{enumerate}
-\item Because we are committed to a fixed syntax, but aim to show that our mechanism does not critically rely on any particular syntactic choices, it is pragmatic to use an existing syntax for which tooling already exists.
-In this paper, we choose Python's syntax.
-Moreover, we use the Python language itself as both the static language and the internal language.
-This makes it easy for us to implement our language, called @\texttt{lang}, itself as a library inside Python, using its quasiquotation and reflection facilities.\footnote{The name was chosen as an initialism for ``actively  typed language'' and because Python marks function decorators, which we use extensively for this purpose, using the @ symbol.} Python is also an interesting starting point  because it  can be thought of as beginning with a degenerate static type system providing just one trivially indexed type constructor, \texttt{dyn} \cite{pfpl}.
+\item We build our mechanism as a library inside Python, \verb|atlang|,  using its quasiquotation and reflection facilities.\footnote{The name was chosen as an initialism for ``actively  typed language'' and because Python marks function decorators, which we use extensively for this purpose, using the @ symbol.} Python is also an interesting starting point  because it  can be thought of as beginning with a  static type system providing just one trivially indexed type constructor, \texttt{dyn} \cite{pfpl}.
+Thus, our mechanism represents a modularly extensible gradual type system.
 Taken together, @\texttt{lang} is almost completely backwards compatible with the Python ecosystem with remarkably little effort: we inherit a large number of  existing tools for working with Python files, Python's substantial packaging infrastructure and access to all existing Python libraries.
-We introduce @\texttt{lang} with practical examples of fragments as libraries in Sec.
+We introduce @\texttt{lang}  in Sec.
 \ref{usage}.
+We continue in Sec.
+\ref{examples} by briefly describing  some more sophisticated examples, including statically typed functional datatypes with nested pattern matching and a complete and statically safe FFI to OpenCL for GPU programming.
 \item In Sec.
@@
 Perhaps surprisingly, we show how a conventional syntax like Python's can be recovered by a purely syntactic desugaring to these term constructors.
-The semantics for @$\lambda$ take the form of a \emph{bidirectionally typed elaboration semantics} where the available type constructors are tracked by a \emph{fragment context}.
+The semantics for @$\lambda$ take the form of a \emph{bidirectionally typed elaboration semantics} where the available type constructors are tracked by a \emph{tycon context}.
 We use a simple lambda calculus as both our static and internal language to emphasize the fundamental character of the approach.
+For more theoretically minded readers or those with no familiarity with Python, this section can be read first.
 \end{enumerate}
@@
 
+\begin{codelisting}[t]
+\begin{lstlisting}
+^\label{imports-start}^from atlib import record, decimal, dyn, string, 
+  string_in, proto, fn, printf^\label{imports-end}^
+
+^\label{hello-ct}^print "Hello, compile-time world!"
+
+^\label{decl-Details}^Details = record[
+  'amount' : decimal[2], 
+  'memo' : dyn]
+^\label{decl-Account-start}^Account = record[
+  'name' : string, 
+  'account_num' : string_in[r'\d{2}-\d{8}']]^\label{decl-Account-end}^
+^\label{decl-Transfer}^Transfer = proto[Details, Account]
+
+^\label{log-transfer-start}^@fn[Transfer, ...]
+def log_transfer(t):
+  """Logs a transfer to the console."""
+  printf("Transferring %s to %s.", 
+    t.amount.to_string, t.name)^\label{log-transfer-end}^
+
+^\label{main-start}^@fn[...]
+def __main__():
+  ^\label{account-start}^account = {name: "Annie Ace", 
+    account_num: "00-00000001"} [:Account]^\label{account-end}^
+  log_transfer(({amount: 5.50, memo: None}, account))
+  log_transfer(({amount: 15.00, memo: "Rent payment"}, 
+    account))^\label{main-end}^
+
+print "Goodbye, compile-time world!"
+\end{lstlisting}
+\caption{[\texttt{listing\ref{example}.py}] An @\texttt{lang} compilation script.}
+\label{example}
+\end{codelisting}
+
 %
 Listing \ref{example} shows an example of a well-typed @\texttt{lang} program for which strong correctness guarantees have been statically maintained using several type constructors defined in libraries.
-As suggested by line \ref{hello-ct}, the top level of every @\texttt{lang} file can be seen as a \emph{compilation script} written directly in Python, the static language.
+As suggested by line \ref{hello-ct}, the top level of every @\texttt{lang} file can be seen as a \emph{compilation script} written directly in Python.
 @\texttt{lang} requires no modifications to the language (version 2.6+ or 3.0+) or features specific to the primary implementation, CPython (so @\texttt{lang} supports alternative implementations like Jython and PyPy).
@@
 %For readers familiar with C/C++, Python can be thought of as serving a role similar to (but more general than) its preprocessor and template system (as we will see).
-\subsection{Active Typechecking}
+%\subsection{Active Typechecking}
+
+The language that the compilation script is written in, Python, is dynamically typed, but functions governed by \verb|atlang| (those annotated with \lstinlinep{@fn} in this example) will be statically typed (Section \ref{s:atc}) and then translated to dynamically typed Python code (Section \ref{s:atr}), both under the control of the type constructors of the types used within them.
+These steps can occur either during a standalone compilation phase (Section \ref{compilation}) or just-in-time upon their first invocation (Section \ref{interactive}).
 
 \paragraph{Types}
-\begin{codelisting}[t]
-\begin{lstlisting}
-^\label{imports-start}^from atlib import record, decimal, dyn, string, 
-  string_in, proto, fn, printf^\label{imports-end}^
-
-^\label{hello-ct}^print "Hello, compile-time world!"
-
-^\label{decl-Details}^Details = record[
-  'amount' : decimal[2], 
-  'memo' : dyn]
-^\label{decl-Account-start}^Account = record[
-  'name' : string, 
-  'account_num' : string_in[r'\d{2}-\d{8}']]^\label{decl-Account-end}^
-^\label{decl-Transfer}^Transfer = proto[Details, Account]
-
-^\label{log-transfer-start}^@fn[Transfer, ...]
-def log_transfer(t):
-  """Logs a transfer to the console."""
-  printf("Transferring %s to %s.", 
-    t.amount.to_string, t.name)^\label{log-transfer-end}^
-
-^\label{main-start}^@fn[...]
-def __main__():
-  ^\label{account-start}^account = {name: "Annie Ace", 
-    account_num: "00-00000001"} [:Account]^\label{account-end}^
-  log_transfer(({amount: 5.50, memo: None}, account))
-  log_transfer(({amount: 15.00, memo: "Rent payment"}, 
-    account))^\label{main-end}^
-
-print "Goodbye, compile-time world!"
-\end{lstlisting}
-\caption{[\texttt{listing\ref{example}.py}] An @\texttt{lang} compilation script.}
-\label{example}
-\end{codelisting}
-Types are constructed programmatically in the static language by applying a type constructor to an index:  \lstinlinep{tycon[idx]}.
+Types are constructed programmatically in the compilation script by applying a type constructor to an index:  \lstinlinep{tycon[idx]}.
 This is in contrast to many statically-typed languages where types (e.g.
-datatypes, interfaces) cannot be constructed computationally, but are only declared ``literally''.
+datatypes, classes) are not constructed programmatically, but can only be declared ``literally''.
 %Put another way,
@@
 \begin{enumerate}
-\item On line \ref{decl-Details}, we construct a functional record type with two (immutable) fields, assigning it to the identifier \lstinlinep{Details}.
+\item On line \ref{decl-Details}, we construct a functional record type with two (immutable) fields, assigning it to the static identifier \lstinlinep{Details}.
 The field \lstinlinep{amount} has type \lstinlinep{decimal[2]}, which classifies decimal numbers having two decimal places, and \lstinlinep{memo} has type \lstinlinep{dyn}, classifying dynamic Python values.
@@
 Syntactically, the index for \lstinlinep{record} is provided using borrowed Python  syntax for array slices (e.g.
-\lstinlinep{a[min:max]}) to approximate conventional notation for type annotations.
+\lstinlinep{a[start:end]}) to approximate conventional notation for type annotations.
 The field names are simply static strings (and could be computed rather than written literally, e.g.
@@
 The field \lstinlinep{name} has type \lstinlinep{string} and the field \lstinlinep{account_num} has a type classifying strings guaranteed statically to be in the regular language specified  by the  regular expression pattern provided as the index (written here as a \emph{raw string literal} \cite{python}).
-This  fragment, based on recent work, statically tracks the language an immutable string is in, providing operations like concatenation, demonstrating the expressive power of our approach \cite{fulton-thesis}.
-%To our knowledge, this type system is not  expressible in other simply typed languages (we do not consider dependently-typed logics without a distinct phase separation).
+This  fragment, based on recent work, statically tracks the language an immutable string is in, providing operations like concatenation, demonstrating the expressive power of our approach \cite{sanitation-psp14}.%To our knowledge, this type system is not  expressible in other simply typed languages (we do not consider dependently-typed logics without a distinct phase separation).
 \item 
@@
 
+\paragraph{Active Type Constructors} Type constructors are implemented in @\texttt{lang} libraries as Python classes inheriting from \lstinlinep{atlang.Type} (classes here serving as Python's form of open sums).
+We show portions of \lstinlinep{atlib.fn} in Listings \ref{fn-top} and \ref{fn-stmts}, \lstinlinep{atlib.record} in Listing \ref{record} and \lstinlinep{atlib.string_in} in Listing \ref{string_in}, which we will discuss as we go on.
+The types in our example are instances of these classes.
+
 \paragraph{Incomplete Types} Incomplete types arise from the application of a type constructor to an index containing an ellipsis (a valid array slice form in Python).
 The static terms \lstinlinep{fn[Transfer, ...]} and \lstinlinep{fn[...]} seen on lines \ref{log-transfer-start} and \ref{main-start} are incomplete types.
-We will discuss how the elided portion of the index (e.g.
-the return type) is determined below.
-
-\paragraph{Active Type Constructors} Type constructors are implemented in @\texttt{lang} libraries as Python classes inheriting from \lstinlinep{atlang.Type} (classes here serving as Python's form of open sums).
-We show portions of \lstinlinep{atlib.fn} in Fig.
-\ref{fn-top}.
-Types are instances of these classes.
-Incomplete types are instead instances of \lstinlinep{atlang.IncompleteType}.
-The compiler controls instantiation by overloading the subscript operator on the \lstinlinep{atlang.Type} class object (using Python's \emph{metaclass} mechanism \cite{python}).
+We will discuss how the elided portions of an index (e.g.
+the return types, which need not be provided) are inferred below.
+
+Incomplete types are not instances of \verb|atlang.Type| but rather instances of \lstinlinep{atlang.IncompleteType}.
+An incomplete type is defined by a type constructor (here \lstinlinep{fn}) and an incomplete index (here, \lstinlinep{(Transfer, Ellipsis)}).
+The language controls instantiation by overloading the subscript operator on the \lstinlinep{atlang.Type} class object (using Python's \emph{metaclass} mechanism \cite{python}).
 
@@
 \end{codelisting}
- To be typechecked by @\texttt{lang} and define run-time behavior, function definitions must have an \emph{ascription}, provided by \emph{decorating} them with either a type or, here, an incomplete type (see \cite{python} for a discussion of Python decorators; we again use Python's metaclasses and operator overloading to support the use of a class as a decorator).
+ To be typechecked by @\texttt{lang} and define run-time behavior, function definitions must have an \emph{ascription}, provided by \emph{decorating} them with either a type or, here, an incomplete type, e.g.
+on lines 14 and 20 (see \cite{python} for a discussion of Python decorators; we again use Python's metaclasses and operator overloading to support the use of a class as a decorator).
+
  Ascribed function definitions do not share Python's semantics and indeed the underlying Python function is discarded immediately after its abstract syntax and static environment (its closure and the global dictionary of the Python module it was defined in) have been extracted by the compiler using Python's built-in reflection capabilities and \lstinlinep{inspect} and \lstinlinep{ast} packages \cite{python}.
 The \lstinlinep{ast} package defines Python's term constructors.
-For function literals (term constructor \lstinlinep{FunctionDef}), the @$\texttt{lang}$ compiler delegates to the type constructor of the ascription in one of two ways, depending on the ascription.
-
-If the ascription is an incomplete type, the compiler invokes the class method \lstinlinep{syn_idx_FunctionDef}, seen on line \ref{syn_idx_FunctionDef}, with the \emph{compiler context} (an object that provides hooks into the compiler and a storage location for accumulating information during typechecking), the incomplete index and the syntax tree of the function definition.
-Here, this method 1) checks and normalizes the incomplete index (here, checking that the argument types are indeed types and turning empty and single arguments into a 0- or 1-tuples for uniformity; not shown); 2) adds the argument names and types to a field of the context that tracks the types of local assignments (initializing it first if it is a top-level function, as in our example); 3) asks the compiler, via the method \lstinlinep{ctx.check}, to typecheck each statement in the body (discussed below); 4) if the term constructor of the last statement is \lstinlinep{ast.Expr} (a top-level expression), then the type of this expression is the return type, otherwise it is \lstinlinep{unit}; and finally, 5) a fully specified index (and thus a type) is synthesized for the function as a whole, here equivalent to \lstinlinep{fn[Transfer, atlib.unit]}.
-Note that this choice of using the last expression as the implicit return value (and considering any statement-level term constructor other than \lstinlinep{Expr} to have a trivial value) is made in a library, not by the language itself.% A different function type constructor could be written to support \lstinlinep{return} statements (omitted for concision).% The choices is made by clients .
-
-Were the ascription a (complete) type, the compiler would delegate to \lstinlinep{fn} by calling an instance method on it, \lstinlinep{ana_FunctionDef}, seen on line \ref{ana_FunctionDef}.
+
+\subsection{Active Typechecking}\label{s:atc}
+We will now trace through how the function \lstinlinep{log_transfer} in Listing 1 is typechecked by \verb|atlang|, via delegation to the type constructors \verb|fn| and \verb|record|, both library constructs.
+
+\paragraph{Functions}
+The @$\texttt{lang}$ compiler begins by delegating to the type constructor of the ascription on the function in one of two ways, depending on whether the ascription is a complete type.
+
+If the ascription is an incomplete type, as in our example, the compiler invokes the class method \lstinlinep{syn_idx_FunctionDef}, seen on line \ref{syn_idx_FunctionDef}, with the \emph{compiler context} (an object that provides hooks into the compiler and a storage location for accumulating information during typechecking), the incomplete index and the syntax tree of the function definition.
+Here, this method \begin{enumerate}
+\item Checks and normalizes the  incomplete function type index (here, \lstinlinep{_check_and_norm_inc} checks that the argument types are indeed types and turns empty and single arguments into a 0- or 1-tuples for uniformity; not shown)
+\item Adds the argument names and types to a field of the context that tracks the types of local assignments (initializing it first if it is a top-level function, as in our example).
+\item Asks the compiler, via the method \lstinlinep{ctx.check}, to typecheck each statement in the body (discussed below)
+\item If the term constructor of the last statement is \lstinlinep{ast.Expr} (a top-level expression), then the type of this expression is the return type, otherwise it is \lstinlinep{atlib.unit}.
+\item A fully specified index is generated, from which a type will be synthesized for the function as a whole, here \lstinlinep{fn[Transfer, atlib.unit]}.
+Note that this choice of using the last expression as the implicit return value (and considering any statement-level term constructor other than \lstinlinep{Expr} to have a trivial value) is made by \verb|atlib|, not by the language itself.
+\end{enumerate}% A different function type constructor could be written to support \lstinlinep{return} statements (omitted for concision).% The choices is made by clients .
+
+Were the ascription a complete type like \verb|fn[Transfer, dyn]|, the compiler would delegate to \lstinlinep{fn} by calling an instance method on it, \lstinlinep{ana_FunctionDef}, seen on line \ref{ana_FunctionDef}, rather than a class method.
 This method proceeds similarly, but does not need to perform the first step and last steps (and does not need to be a class method) because the index was already determined when the type was constructed, so it can be accessed via \lstinlinep{self.idx}.
-If the final statement is an expression, is analyzed against the provided return type (see below).
-Otherwise, only \lstinlinep{unit} or \lstinlinep{dyn} are valid return types (again, a choice made in a library)%.
+If the final statement is an expression, it is analyzed against the provided return type (see below).
+Otherwise, only \lstinlinep{unit} or \lstinlinep{dyn} are valid return types (again, a choice made by \verb|atlib|).%.
 If it is, the expression .
@@
 \end{codelisting}
-The \lstinlinep{ctx.check} method is defined by the compiler.
-As is the pattern in this paper, the compiler delegates to an active type constructor based on the term constructor of the statement being checked.
-Most statement-level term constructors other than \lstinlinep{Expr} simply delegate to the type constructor of the function they are defined within by calling class methods of the form \lstinlinep{check_TermCon}, where \lstinlinep{TermCon} is a term constructor or combination of term constructors in a few cases where a finer distinction than what Python itself made was necessary, as defined by \lstinlinep{ast}.
+The \lstinlinep{ctx.check} method mentioned above is defined by the compiler.
+As is the pattern, it simply delegates to an active type constructor based on the term constructor of the statement being checked.
+Most statement-level term constructors other than \lstinlinep{Expr} simply delegate to the type constructor of the function they are defined within by calling class methods of the form \lstinlinep{check_TermCon}, where \lstinlinep{TermCon} is a term constructor or combination of term constructors in a few cases where a finer distinction than what Python itself made was necessary.
 
@@
 
-\paragraph{Bidirectional Typechecking of Expressions} The methods \lstinlinep{ctx.ana} and \lstinlinep{ctx.syn} are also defined by the compiler and can be called by type constructors to request type analysis (when the expected type is known) and synthesis (when the type is an ``output''), respectively, for an expression.
+\paragraph{Bidirectional Typechecking} The methods \lstinlinep{ctx.ana} and \lstinlinep{ctx.syn} are also defined by the compiler and can be called by type constructors to request type analysis (when the expected type is known) and synthesis (when the type is an ``output''), respectively, for an expression.
 Once again, the compiler delegates to a type constructor by a protocol that depends on the term constructor, invoking methods of the form \lstinlinep{ana_TermCon} or \lstinlinep{syn_TermCon}.
@@
 Two types with different type constructors are never equal, to ensure that the burden of ensuring that typing respects type equality is local to a single type constructor.
-No form of subtyping or implicit coercion is supported for the purposes of this paper (though we have designed a mechanism that similarly localizes reasoning to a single type constructor, we do not have room to precisely describe it here.)
+No form of subtyping or implicit coercion is supported for the purposes of this paper (though we have designed a mechanism that similarly localizes reasoning to a single type constructor, we do not  describe it here.)
 
@@
 
-\subsection{Active Translation}
+\subsection{Active Translation}\label{s:atr}
 Once typechecking is complete, the compiler enters the translation phase.
@@
 
-\subsection{Interactive Invocation} @\texttt{lang} functions over values of type \lstinlinep{dyn} can be invoked directly from Python.
+\subsection{Interactive Invocation}\label{interactive} @\texttt{lang} functions over values of type \lstinlinep{dyn} can be invoked directly from Python.
 Typechecking and translation occurs upon first invocation (subsequent calls are cached; we assume that the static environment is immutable).
@@
 %Because \lstinlinep{t} will be statically guaranteed to have type \lstinlinep{Transfer},
+
+\section{More Examples}\label{examples}
+In the previous section, we showed examples of several interesting type system fragments implemented as libraries using \verb|atlang|, including functional record types, a prototypic object system and regular string types.
+A more detailed description of regular string types and their implementation using \verb|atlang| was recently published \cite{sanitation-psp14}.
+Here, we showcase two more powerful examples that demonstrate the expressive power of system: functional datatypes with nested pattern matching, and a type safe foreign function interface to the complete OpenCL language for many-core programming on devices like GPUs.
+
+\subsection{Functional Datatypes and Nested Pattern Matching}
+A powerful feature of modern functional languages like ML is support for nested pattern matching over datatypes (i.e.
+recursive labeled disjoint sums) and tuples.
+For example, tree structures are well-modeled in this way, as we show in Listing \ref{example-casetypes}.
+Here, \verb|Tree| is a \emph{case type}, which is what we call functional datatypes in \verb|atlib|.
+The type constructor \verb|Tree| is declared as a named type using the \verb|@ty| annotation.
+This is a general mechanism in \verb|atlang| for supporting recursive types that behave generatively, i.e.
+that are identified by a name, rather than a structure.
+Functional datatypes have the same character.
+The \verb|@ty| annotation also supports parameterized types.
+Here, \verb|a| must be another type.
+The type index is a series of cases with, optionally, corresponding types.
+Here, a (binary) tree can either be empty (no associated data), a leaf with an associated value of type \verb|a|, or an internal node, which takes a pair of trees as data.
+We use the \verb|tuple| type constructor to represent pairs.
+
+The function \verb|treesum| on lines 11-17 takes a tree containing \verb|dyn| values and computes the sum of these values.
+The case operator is defined by the \verb|casetype| type constructor, leveraging a combination of the \verb|Attribute| and the \verb|Subscript|  term constructors (both are targeted operations, see above).
+The slice syntax is again re-interpretted, here s a series of rules, each of which starts with a pattern, followed by a colon, then a case.
+Variables in patterns can be bound in the case.
+Patterns can consist of case names and tuple patterns at arbitrary depth.
+More generally, we have done initial designs for an extensible pattern language, again delegating to type constructors to determine whether the pattern is valid and which variables it binds, but here we restrict ourselves to these two type constructors.
+
+By combining the functional behavior of the \verb|fn| type constructor, which dispenses with ``return'' statements, and creatively repurposing existing syntax, we are thus able to implement an essentially idiomatic statically-typed functional language, entirely as a library for Python.
+The constructs we defined can be composed arbitrarily with, e.g.
+the record types or regular string types in the previous section, without any concern regarding syntactic or semantic conflict, because \verb|casetype| only controls the semantics of the \verb|case| operation on values that are of such a type.
+
+\subsection{A Low-Level Foreign Function Interface to OpenCL} 
+\begin{codelisting}[t]
+\begin{lstlisting}
+from atlang import ty
+from atlib import casetype, tuple, fn
+
+@ty
+def Tree(a): casetype[
+  "Empty",
+  "Leaf" : a,
+  "Node": tuple[Tree(a), Tree(a)]
+]
+
+@fn
+def treesum(tree : Tree(dyn)) -> dyn:
+  tree.case[
+    Empty: 0,
+    Leaf(x): x,
+    Node((l, r)): treesum(l) + treesum(r)
+  ]
+\end{lstlisting}
+\caption{An example of case types and nested pattern matching.}
+\label{example-casetypes}
+\end{codelisting}
+
+\begin{codelisting}[t]
+\begin{lstlisting}
+import atlib.opencl as opencl
+
+dev = opencl.init_device(0)
+buffer = dev.send(numpy.zeros((80000,)))
+buffer_t = buffer.ty # == opencl.global_ptr[elem_t]
+elem_t = buffer_t.elem_t # == opencl.double
+
+@opencl.fn
+def map(x : buffer_t, f : opencl.fn[elem_t, elem_t]):
+  gid = get_global_id(0)
+  x[gid] = f(x[gid])
+
+@opencl.fn
+def mad(x : elem_t, m : elem_t, y : elem_t):
+  return m*x + y
+@opencl.fn
+def my_mad(x : elem_t): return mad(x, 10, 5)
+@opencl.fn
+def my_mad2(x : elem_t): return mad(x, 20, 5)
+
+map(buffer, my_mad)
+map(buffer, my_mad2)
+\end{lstlisting}
+\caption{An example of a type system for OpenCL.}
+\label{example-opencl}
+\end{codelisting}
+
+\begin{codelisting}[t]
+\begin{lstlisting}
+#pragma OPENCL EXTENSION cl_khr_fp64 : enable
+
+double mad(double x, double m, double y) {
+  return m*x + y;
+}
+
+double my_mad(double x) {
+  return mad(x, 10.0, 5.0);
+}
+
+double my_mad2(double x) {
+  return mad(x, 20.0, 5.0);
+}
+
+kernel void map__1_(double x) {
+  size_t gid = get_global_id(0);
+  x[gid] = my_mad(x[gid]);
+}
+
+kernel void map__2_(double x) {
+  size_t gid = get_global_id(0);
+  x[gid] = my_mad2(x[gid]);
+}
+\end{lstlisting}
+\caption{The underlying code generated by atlib.opencl.}
+\label{example-opencl-out}
+\end{codelisting}
+
+Python is a common language in scientific computing and data analysis domains.
+The performance of large-scale analyses can often be a bottleneck, so programmers often wish to write and interact with code written in a low-level language.
+OpenCL is designed precisely for this workflow, exposing a C-based language that can compile to a variety of specialized hardware (e.g.
+GPUs) via a standard API.
+The \verb|pyopencl| package exposes this API to Python code and integrates it with the popular \verb|numpy|numeric array package \cite{klockner2011pycuda}.
+To compile an OpenCL function, however, users must write it as a string.
+This is neither idiomatic nor safe, because it defers typechecking to invocation-time.
+
+In Listing \ref{example-opencl}, we show usage of the \verb|atlib.opencl| package, which implements the entirety of the OpenCL language (which includes essentially the entirety of C99, plus some extensions to work with multiple memory spaces).
+Although the details are beyond the scope of this paper, we note that the mapping onto Python syntax was straightforward, particularly given the flexibility of analytic numeric literal forms, as described above.
+
+After initializing a device, line 4 sends a \verb|numpy| array to the device, assigning \verb|buffer| its handle.
+Line 5 extracts its type, which is based on the dynamic element type of the \verb|numpy| array.
+The type constructor \verb|global_ptr| defines the semantics of OpenCL's pointers to global memory (e.g.
+subscripting, pointer arithmetic, pointer differencing), consistent with the C99 specification.
+Because the compilation script is programmatic, we can extract its element type programmatically as well.
+Here, it is the \verb|opencl.double| type, which supports arithmetic operations, including the full array of widening rules defined by C99 when paired with any of the other numeric OpenCL types (using handle sets, above, to prioritize control to the wider type).
+
+Lines 8-19 define several OpenCL functions using Python syntax, but retaining OpenCL's semantics.
+The \verb|map| function takes a buffer argument and a function over its elements and applies the function to each element in the buffer in a data parallel manner.
+Each thread handles the element corresponding to its thread ID, called a \verb|group ID| in OpenCL and accessed by a primitive on the first line.
+Note that OpenCL does not itself have support for higher-order functions, however.
+Instead, higher order functions are essentially an extension to the foreign language being encoded using active type constructors.
+The OpenCL function decorator treats uses of higher order functions as if each function ever provided to it had a \emph{singleton type}, i.e.
+a type inhabited only by that particular function.
+Each call lazily generates code for a specialized variant of that function.
+The result can be seen in Listing \ref{example-opencl-out}, which shows the OpenCL code generated (as a string sent through \verb|pyopencl|) by this example.
+
+Though we do not show any other extensions atop the OpenCL library here, it is straightforward to implement variants of those described above that target their translation phase to OpenCL rather than Python.
+In many cases, the typechecking code can be inherited directly.
+This is, we argue, rather compelling: a decidedly low-level language like OpenCL can be extended with low-overhead versions of sophisticated features, like a prototypic object system, modularly, via its foreign function interface from a scripting language like Python.
 
@@
 \paragraph{Fragment Client Perspective} A program, $\rho$, consists of a series of fragment imports, $\Phi$, defining active type constructors for use in an expression, $e$.
-As described in the introduction, expressions are indexed by static terms, $\sigma$, and indices that can only be trivial are omitted.
+Expression forms can be indexed by static terms, $\sigma$.
 The abstract syntax of $e$ provides $\FF{let}$ binding of variables and for convenience, static values can also be bound to a static variable, $\tvar{x}$ (distinguished in bold for clarity), using $\FF{slet}$.
@@
 \paragraph{Metatheory} The formulation shown here guarantees that type synthesis actually produces a type, given well-formed contexts.
-The definitions are straightforward and the proof is a simple induction with simple lemmas about the normalization semantics.
+The definitions are straightforward and the proof is a simple induction.
+We write $\fvalCtx~\mathtt{fragment}$ for fragments with no duplicate tycon names and closed tycon definitions only, and $\Gamma~\mathtt{ctx}_\fvalCtx$ for typing contexts that only map variables to types constructed with tycons in $\fvalCtx$.
 %We plan to prove the full proofs in a companion tech report (available upon request).
@@
 
-We can also prove that importing additional type constructors cannot change the semantics of a previously well-typed term, assuming that naming conflicts have been resolved by some extrinsic mechanism.
+We also have that importing additional type constructors cannot change the semantics of a previously well-typed term, assuming that naming conflicts have been resolved by some extrinsic mechanism.
 Indeed, the proof is an essentially trivial induction because of the way we have structured our mechanism.
 The type constructor delegated responsibility over a term is deterministically determined irrespective of the structure of $\fvalCtx$.
-We plan to provide full proofs of these theorems in a technical report.
-
-\begin{theorem}[Unambiguous Extension]
+%We plan to provide full proofs of these theorems in a technical report.
+
+\begin{theorem}[Stable Extension]
 If $\fvalCtx~\mathtt{fragment}$ and $\Gamma~\mathtt{ctx}_\fvalCtx$ and $\atjsynX{e}{\sigma}{\iota}$ and $\fvalCtx'~\mathtt{fragment}$ and $\text{dom}(\fvalCtx) \cap \text{dom}(\fvalCtx') = \emptyset$ then ${\Gamma}\vdash_{\fvalCtx;\fvalCtx'}{e}\Rightarrow{\sigma}\leadsto{\iota}$.
 \end{theorem}
-%Metatheoretic guarantees about the translation cannot be provided in the formulation as given.
-However, inserting straightforward checks to guarantee that the translation is a closed term, and provide simple mechanisms for hygiene, would be simple, but are omitted.
+
+Other metatheoretic guarantees about the translation cannot be provided in the formulation as given.
+However, inserting straightforward checks to guarantee that the translation is a closed term, and provide simple mechanisms for hygiene, would be simple, but are omitted to keep our focus on the basic structure of the calculus as a descriptive artifact.
 %\noindent
@@
 
-Although we do not have space to detail it here, we have conducted a substantial case study using (an older version of) @\texttt{lang}: an implementation of the entirety of the OpenCL type system (a variant of C99), as well as several extensions to it, as a library.
+Although we only touched on the details here, we have conducted a substantial case study using @\texttt{lang}: an implementation of the entirety of the OpenCL type system (a variant of C99), as well as several extensions to it, as a library.
 Our operations translate to Python's lower-level FFI with OpenCL but guarantee type safety at the interface between languages.
-We have built a neurobiological circuit simulation framework atop this library as well.
-Based on our examples in this paper and this admittedly anecdotal experience report, we submit that active type constructors deserve broad consideration by the research community as a foundational technique for structuring typed programming languages.
+A neurobiological circuit simulation framework has been built atop this library (a detailed case study is in preparation).
+%Based on our examples in this paper and this admittedly anecdotal experience report, we submit that active type constructors deserve broad consideration by the research community as a foundational technique for structuring typed programming languages.
 We are currently refactoring these projects against the (new) bidirectional semantics described here.

### 2014-10-14 atlang-modularity15/atlang-modularity15.tex
updated submission to modularity due to extension

@@
   basicstyle=\ttfamily\scriptsize,
-  morekeywords={lambda, self, assert, cls, @classmethod, with},
-  deletekeywords={min,max},
+  morekeywords={lambda, self, assert, cls, @classmethod, with,as},
+  deletendkeywords={tuple,id,buffer,map},
   numbers=left,
@@
 
-\newcommand{\lstinlinep}[1]{\lstinline[language=Python,basicstyle=\ttfamily\small]{#1}}
+\newcommand{\lstinlinep}[1]{\lstinline[language=Python,basicstyle=\ttfamily\small,deletendkeywords={tuple,buffer,map}]{#1}}
 
@@
 Programmers prefer using abstractions available as libraries for the language they already use over those requiring a switch to a new language.
-But library providers cannot typically extend the type systems of these languages directly, so powerful statically typed abstractions  remain obscure.
-This suggests that a language is needed that is {backwards compatible} with the libraries and tools of a widely adopted language, but that has an  {internally extensible type system}, organized around a mechanism that guarantees that any combination of extensions can be imported safely.
+But library providers cannot typically extend the static type systems of these languages directly, so powerful statically typed abstractions  remain obscure.
+This suggests the need for a mechanism that allows abstraction providers to extend the type system of a widely adopted language from within, in a way that guarantees that any combination of extensions can be used together without worrying about conflicts.
 
 Our aim in this paper is to 1) introduce such a mechanism; and 2) show how it can be implemented as a library for an existing widely used language, here Python.
-We describe two artifacts: @\texttt{lang}, an  extensible statically typed language implemented as a library inside Python, and @$\lambda$, a minimal core calculus.
-Both @\texttt{lang} and @$\lambda$ have a fixed syntax but  determine the semantics of each term by delegating control, according to a simple but powerful bidirectionally typed elaboration semantics, to methods associated with a relevant \emph{active type constructor}, defined in a library.
-This process is stable even when additional extensions are imported, so we avoid ambiguity by construction.
-We describe several examples of interesting type system fragments as libraries, including functional constructs, a type system for secure string sanitation and a foreign function interface to OpenCL.% By flexibly  repurposing a fixed syntax in this type-directed manner, we avoid ambiguities by construction.
+We describe two artifacts: @\texttt{lang}, an  extensible statically typed language implemented as a library within Python, and @$\lambda$, a minimal core calculus.
+Both @\texttt{lang} and @$\lambda$ have a fixed syntax.
+They determine the semantics of each term by delegating control, according to a simple but powerful bidirectionally typed elaboration semantics, to methods associated with a relevant \emph{active type constructor}, defined in a library.
+This process is stable even when additional extensions are imported, so we avoid syntactic and semantic ambiguity by construction.
+We validate our approach with several examples of interesting type system fragments that are built into other languages but can be implemented as libraries using our approach, including functional constructs, a type system for secure string sanitation and a typed foreign function interface to OpenCL.% By flexibly  repurposing a fixed syntax in this type-directed manner, we avoid ambiguities by construction.
 %In @\texttt{lang}, types and functions are defined using Python's syntax, distributed using Python's package system and can call into Python libraries, so @\texttt{lang} is uniquely backwards compatible.
@@
 In these situations, abstraction providers generally develop a new language dialect, increasingly with the aid of tools like language workbenches \cite{erdweg2013state}.
-Unfortunately, this \emph{language-oriented approach} \cite{journals/stp/Ward94} is problematic at language boundaries: working with components written in several foreign languages often requires programming against a low-level internal implementation, rather than its semantics, via a foreign-function interface (FFI).
+Unfortunately, this \emph{language-oriented approach} \cite{journals/stp/Ward94} is problematic at language boundaries: working with components written in several distinct languages often requires programming against their low-level implementation details, rather than their semantics, via a foreign-function interface (FFI) to a common implementation language (e.g.
+C or the JVM).
 This is brittle, unnatural and unsafe and can cause performance issues due to necessary checks and conversions at language boundaries.
-As more languages are used together,  more FFIs must be developed and traversed.%To avoid this, it is necessary to combine the different languages and their associated tools, but this often requires significant effort and expertise (we will discuss this further in Sec.
+As more languages are used together, more FFIs need to be constructed and traversed.%To avoid this, it is necessary to combine the different languages and their associated tools, but this often requires significant effort and expertise (we will discuss this further in Sec.
 \ref{related-work}).
 
-These issues can prevent even dialects introducing only a few new constructs from being adopted.
+These issues prevent even those dialects that introduce only a few new constructs from being adopted.
 A study comparing a Java dialect, Habanero-Java (HJ), with a comparable library, \lstinlinep{java.util.concurrent}, for example, found that the language-based abstractions in HJ were easier to use and provided useful static guarantees \cite{cave2010comparing}.
@@
 
-Extensible languages have promised to reduce the problems associated with distinct dialects by giving abstraction providers more direct control over a language's syntax and semantics and supporting the composition of these language fragments within or alongside libraries.
+Extensible languages have promised to reduce the problems associated with distinct dialects by giving abstraction providers more direct and granular control over a language's syntax and semantics from within, or alongside, libraries.
 %In   an internally extensible language, programmers  can \emph{granularly} import the primitive constructs best suited to each component of their application or library.
@@
 
-Our focus in this paper is to enable the introduction of new static type system fragments as safely composable libraries for an existing, widely-used language.
-The most immediately difficult problems related to composing type system fragments arise when attempting to enable the decentralized introduction of new concrete syntax or corresponding term constructors to support the operations of this new type system.
-Syntactic ambiguities are difficult to modularly preclude (though some techniques exist).
+Our focus in this paper is to enable the introduction of new static type system fragments as safely composable libraries for an existing and widely-used language, leaving its concrete and abstract syntax fixed.
+This choice is freeing: questions of syntactic ambiguity no longer need to be addressed.
 Moreover, many tools operate by exhaustive case analysis over the concrete or abstract syntax of the terms of the language they work with (e.g.
 syntax highlighters, editor modes, pretty-printers, style checkers, documentation generators).
-Indeed, enabling the decentralized introduction of new term constructors is the standard example used when describing the more general problem of enabling the extension of a sum type with new cases as well as new  functions over its cases  \cite{Reynolds75}, prompting Wadler to dub it the \emph{expression problem} \cite{wadler1998expression}.
-
-\paragraph{Approach} Fewer tools operate by exhaustive case analysis over type constructors.
-In a conventionally formulated programming language, however, it is difficult to introduce a new type system fragment without adding new concrete syntax or term constructors to accompany the new type constructors.
-In this paper, we aim to show that this need not be the case.
-We introduce a method that allows library providers to introduce new type system fragments by defining only new type constructors, in libraries, while leaving the concrete syntax, the term constructors and the target language of compilation fixed.
-The key idea is that we determine the semantics of each term not based directly on the identity of its term constructor (the usual ``syntax-directed'' approach), but rather by delegating to a statically evaluated function associated with a relevant type constructor.
+These tools will continue to work.
+%Indeed, enabling the decentralized introduction of new term constructors is the standard example used when describing the more general problem of enabling the extension of a sum type with new cases as well as new  functions over its cases  \cite{Reynolds75}, prompting Wadler to dub it the \emph{expression problem} \cite{wadler1998expression}.
+
+\paragraph{Approach} Our key insight is that few tools operate by exhaustive case analysis over type constructors.
+We introduce a mechanism that allows library providers to introduce new type system fragments by defining only new type constructors, in libraries, while leaving the concrete and abstract syntax, and the base semantics of the language, fixed.
+
+To support new operations, the semantics of terms are no longer determined directly based on the identity of their term constructors (the usual ``syntax-directed'' approach).
+Instead, we delegate to a statically defined function associated with a relevant type constructor.
 The term constructor's identity only serves to determine how the delegate is assigned.
-We call user-defined type constructors equipped with static functions and distributed in libraries \emph{active type constructors}.
-
-As a simple example, consider the concrete term \lstinlinep{e.r} (where \lstinlinep{e} is an expression and \lstinlinep{r} is a static label).
-A fragment defining statically typed pairs might want this to refer to projection of the right component of the pair, while a fragment defining functional record types might want it project out the field labeled \verb|r| and an object system might introduce a more complex semantics involving dynamic dispatch.
-Were we to naively ``open up'' the syntax and semantics of the language, these fragments would conflict.
-
-Instead, we assume that the syntax is fixed and includes this commonly used form, mapping it to abstract term  $\FF{attr}[\text{``r''}](e)$ (i.e.
+We call user-defined type constructors equipped with these static functions and distributed in libraries \emph{active type constructors}.
+
+As a simple example, consider the concrete term \lstinlinep{e.r} (where \lstinlinep{e} is an expression and \lstinlinep{r} is a static label), which maps to the abstract term $\FF{attr}[\text{``r''}](e)$ (i.e.
 the term constructor $\FF{attr}$, \emph{indexed} by a static representation of the label and taking one \emph{argument}, $e$).
-Our semantics delegate responsibility for this term to a static function associated with the type constructor of the type recursively synthesized by the ``target'' of the operation, $e$.
-Tuples, records, objects and variants thereof can all be supported over a common syntax.
+A fragment defining statically typed pairs might see this term as referring to projection of the right component of the pair $e$, while a fragment defining functional record types might want it project out a field labeled \verb|r| from $e$, and an object system might introduce a more complex semantics involving dynamic dispatch.
+The type of dynamically classified Python values would follow Python's own complex field lookup protocol.
+Were we to naively ``open up'' the syntax and semantics, these fragments would conflict.
+
+Instead, our semantics will delegate responsibility for assigning a type and a translation to this term to a static function associated with the type constructor of the type recursively synthesized by $e$, the \emph{target} of the operation.
+Thus, tuples, records, objects, dynamic values and variants thereof can  share this syntax.
+
 As we will detail, terms for which there is no clear ``target'' from which to determine a delegate (e.g.
-number or list literals) instead either use the type they are being analyzed against (consistent with how extensible parsing is done in Wyvern \cite{TSLs}) or, if not available, a ``default'' delegate determined by the type constructor governing the function in which they appear.
+number or list literals) instead either use the type they are being analyzed against (following the approach for composable syntax extensions as supported in Wyvern \cite{TSLs}) or, if not available, a ``default'' delegate determined by the type constructor governing the function in which they appear.
 Symmetric binary terms (e.g.
 addition) use a third protocol, which we will discuss.
-There is always exactly one active type constructor delegated  responsibility over a term and this delegate is stable with respect to additional extension imports, so semantic ambiguities cannot arise by construction.
+There is always exactly one active type constructor delegated  responsibility over a term and this delegate is stable with respect to additional extension imports, so semantic ambiguities cannot arise by construction (unlike systems that treat the type system as a ``bag of rules'').
 %This is a generalization of a previous mechanism that we have developed, which deals only with extending the elaboration, but not the typechecking, of the bodies of delimited literals \cite{omar-ecoop14}.
@@
 %
-Listing \ref{example} shows an example of a well-typed @\texttt{lang} program for which strong correctness guarantees have been statically maintained using several type constructors defined in libraries.
+Listing \ref{example} shows an example of a well-typed @\texttt{lang} program for which stronger correctness guarantees than Python can statically provide have been established using several type system fragments defined in libraries.
 As suggested by line \ref{hello-ct}, the top level of every @\texttt{lang} file can be seen as a \emph{compilation script} written directly in Python.
-@\texttt{lang} requires no modifications to the language (version 2.6+ or 3.0+) or features specific to the primary implementation, CPython (so @\texttt{lang} supports alternative implementations like Jython and PyPy).
+@\texttt{lang} requires no modifications to the language (version 2.6+ or 3.0+) or features specific to its main implementation, CPython (so @\texttt{lang} supports alternative implementations like Jython and PyPy).
 This choice pays off immediately on the first line: @\texttt{lang}'s import mechanism is Python's import mechanism, so Python's build tools (e.g.
 \lstinlinep{pip}) and package repostories (e.g.
-\lstinlinep{PyPI}) are directly available for distributing @\texttt{lang} libraries, including those defining type constructors.
+\lstinlinep{PyPI}) are directly available for distributing @\texttt{lang} libraries defining type system fragments.
 Here, \lstinlinep{atlib} is the ``standard library'' but does not benefit from special support in \lstinlinep{atlang} itself.% Nothing in \lstinlinep{atlib} benefits from special support from \lstinlinep{atlang}, the core language package.
@@
 \paragraph{Types}
-Types are constructed programmatically in the compilation script by applying a type constructor to an index:  \lstinlinep{tycon[idx]}.
-This is in contrast to many statically-typed languages where types (e.g.
-datatypes, classes) are not constructed programmatically, but can only be declared ``literally''.
+Types are constructed programmatically in the compilation script, always taking the form of a type constructor applied to a type index:  \lstinlinep{tycon[idx]}.
+Unlike many statically-typed languages, types are constructed programmatically, rather than  declared ``literally'' (e.g.
+classes in Java, datatypes in ML).
 %Put another way,
@@
 \begin{enumerate}
-\item On line \ref{decl-Details}, we construct a functional record type with two (immutable) fields, assigning it to the static identifier \lstinlinep{Details}.
-The field \lstinlinep{amount} has type \lstinlinep{decimal[2]}, which classifies decimal numbers having two decimal places, and \lstinlinep{memo} has type \lstinlinep{dyn}, classifying dynamic Python values.
+\item On line \ref{decl-Details}, we construct a functional record type with two (immutable) fields, naming it (by assigning it to a static identifier) \lstinlinep{Details}.
+The field \lstinlinep{amount} has type \lstinlinep{decimal[2]}, which classifies decimal numbers having two decimal places, and \lstinlinep{memo} has type \lstinlinep{dyn}, which comes already constructed, classifying dynamic Python values.
 %Here, \lstinlinep{record} and \lstinlinep{decimal} are {indexed type constructors}.
-Syntactically, the index for \lstinlinep{record} is provided using borrowed Python  syntax for array slices (e.g.
+For convenience, the index for a \lstinlinep{record} type can be provided using borrowed Python  syntax for array slices (e.g.
 \lstinlinep{a[start:end]}) to approximate conventional notation for type annotations.
@@
 The field \lstinlinep{name} has type \lstinlinep{string} and the field \lstinlinep{account_num} has a type classifying strings guaranteed statically to be in the regular language specified  by the  regular expression pattern provided as the index (written here as a \emph{raw string literal} \cite{python}).
-This  fragment, based on recent work, statically tracks the language an immutable string is in, providing operations like concatenation, demonstrating the expressive power of our approach \cite{sanitation-psp14}.%To our knowledge, this type system is not  expressible in other simply typed languages (we do not consider dependently-typed logics without a distinct phase separation).
+This  fragment, based on recent work, statically tracks the regular language an immutable string is a member of through operations like concatenation and substitution, which is useful for ensuring well-formedness and preventing injection attacks \cite{sanitation-psp14}.%To our knowledge, this type system is not  expressible in other simply typed languages (we do not consider dependently-typed logics without a distinct phase separation).
 \item 
@@
 
-\paragraph{Active Type Constructors} Type constructors are implemented in @\texttt{lang} libraries as Python classes inheriting from \lstinlinep{atlang.Type} (classes here serving as Python's form of open sums).
+\paragraph{Active Type Constructors} Type constructors are Python classes inheriting from \lstinlinep{atlang.Type} (classes here serving as Python's form of open sums).
 We show portions of \lstinlinep{atlib.fn} in Listings \ref{fn-top} and \ref{fn-stmts}, \lstinlinep{atlib.record} in Listing \ref{record} and \lstinlinep{atlib.string_in} in Listing \ref{string_in}, which we will discuss as we go on.
@@
 We will discuss how the elided portions of an index (e.g.
-the return types, which need not be provided) are inferred below.
-
-Incomplete types are not instances of \verb|atlang.Type| but rather instances of \lstinlinep{atlang.IncompleteType}.
+the return types, which need not be provided) will be inferred below.
+
+Incomplete types are not instances of \lstinlinep{atlang.Type} but rather instances of \lstinlinep{atlang.IncompleteType}.
 An incomplete type is defined by a type constructor (here \lstinlinep{fn}) and an incomplete index (here, \lstinlinep{(Transfer, Ellipsis)}).
-The language controls instantiation by overloading the subscript operator on the \lstinlinep{atlang.Type} class object (using Python's \emph{metaclass} mechanism \cite{python}).
+The language controls whether the form \lstinlinep{tycon[idx]} results in a complte or incomplete type by overloading the subscript operator on the \lstinlinep{atlang.Type} class object (using Python's \emph{metaclass} mechanism \cite{python}).
 
@@
     else: rty = unit
-    return (arg_types, rty) # fully specified index
+    return (arg_types, rty) # complete index
 
@@
  Ascribed function definitions do not share Python's semantics and indeed the underlying Python function is discarded immediately after its abstract syntax and static environment (its closure and the global dictionary of the Python module it was defined in) have been extracted by the compiler using Python's built-in reflection capabilities and \lstinlinep{inspect} and \lstinlinep{ast} packages \cite{python}.
-The \lstinlinep{ast} package defines Python's term constructors.
+The \lstinlinep{ast} package defines Python's term constructors and we will refer to these throughout the paper.
 
 \subsection{Active Typechecking}\label{s:atc}
-We will now trace through how the function \lstinlinep{log_transfer} in Listing 1 is typechecked by \verb|atlang|, via delegation to the type constructors \verb|fn| and \verb|record|, both library constructs.
+We will now trace through how the functions \lstinlinep{log_transfer} and \lstinlinep{__main__} in Listing 1 are typechecked by \verb|atlang|, via delegation to the type constructors just mentioned.
 
 \paragraph{Functions}
-The @$\texttt{lang}$ compiler begins by delegating to the type constructor of the ascription on the function in one of two ways, depending on whether the ascription is a complete type.
-
-If the ascription is an incomplete type, as in our example, the compiler invokes the class method \lstinlinep{syn_idx_FunctionDef}, seen on line \ref{syn_idx_FunctionDef}, with the \emph{compiler context} (an object that provides hooks into the compiler and a storage location for accumulating information during typechecking), the incomplete index and the syntax tree of the function definition.
-Here, this method \begin{enumerate}
-\item Checks and normalizes the  incomplete function type index (here, \lstinlinep{_check_and_norm_inc} checks that the argument types are indeed types and turns empty and single arguments into a 0- or 1-tuples for uniformity; not shown)
+The @$\texttt{lang}$ compiler begins by delegating to the type constructor of the ascription on the function in one of two ways, depending on whether the ascription is complete.
+
+If the ascription is an incomplete type, as in our example, the compiler invokes the class method \lstinlinep{syn_idx_FunctionDef}, seen on line \ref{syn_idx_FunctionDef}, passing it the \emph{compiler context} (an object that provides hooks into the compiler and a storage location for accumulating information during typechecking), the incomplete type index and the syntax tree of the function definition.
+Here, this method: \begin{enumerate}
+\item Checks and normalizes the  incomplete function type index.
+\lstinlinep{_check_and_norm_inc} (not shown) checks that the argument types are indeed types and turns empty and single arguments into a 0- or 1-tuples for uniformity.
 \item Adds the argument names and types to a field of the context that tracks the types of local assignments (initializing it first if it is a top-level function, as in our example).
 \item Asks the compiler, via the method \lstinlinep{ctx.check}, to typecheck each statement in the body (discussed below)
-\item If the term constructor of the last statement is \lstinlinep{ast.Expr} (a top-level expression), then the type of this expression is the return type, otherwise it is \lstinlinep{atlib.unit}.
-\item A fully specified index is generated, from which a type will be synthesized for the function as a whole, here \lstinlinep{fn[Transfer, atlib.unit]}.
+\item If the term constructor of the last statement is \lstinlinep{ast.Expr} (i.e.
+a top-level expression), then the type of this expression is the return type, otherwise it is \lstinlinep{atlib.unit}.
 Note that this choice of using the last expression as the implicit return value (and considering any statement-level term constructor other than \lstinlinep{Expr} to have a trivial value) is made by \verb|atlib|, not by the language itself.
+\item A complete index is generated, from which a type will be synthesized for the function, here \lstinlinep{fn[Transfer, atlib.unit]} because uses of the \lstinlinep{printf} operator, also defined in \lstinlinep{atlib}, have type \verb|atlib.unit| (i.e.
+\lstinlinep{printf} is used only for its effect).
 \end{enumerate}% A different function type constructor could be written to support \lstinlinep{return} statements (omitted for concision).% The choices is made by clients .
 
-Were the ascription a complete type like \verb|fn[Transfer, dyn]|, the compiler would delegate to \lstinlinep{fn} by calling an instance method on it, \lstinlinep{ana_FunctionDef}, seen on line \ref{ana_FunctionDef}, rather than a class method.
+Were the ascription a complete type like \lstinlinep{fn[Transfer, dyn]}, the compiler would delegate to \lstinlinep{fn} by calling an instance method on it, \lstinlinep{ana_FunctionDef}, seen on line \ref{ana_FunctionDef}, rather than a class method.
 This method proceeds similarly, but does not need to perform the first step and last steps (and does not need to be a class method) because the index was already determined when the type was constructed, so it can be accessed via \lstinlinep{self.idx}.
-If the final statement is an expression, it is analyzed against the provided return type (see below).
-Otherwise, only \lstinlinep{unit} or \lstinlinep{dyn} are valid return types (again, a choice made by \verb|atlib|).%.
+If the final statement is an expression, it is analyzed against the provided return type (see below), rather than asked to synthesize a type.
+Functions that don't end in expressions can only have return types \lstinlinep{unit} or \lstinlinep{dyn} (again, a choice made by \verb|atlib|).%.
 If it is, the expression .
@@
     except KeyError:
-      try: return self._syn_lift(ctx.static_env[e.id])
+      try: return ctx.syn_Lift(ctx.static_env[e.id])
       except KeyError: raise atlang.TypeError("...", e)
@@
 
-\paragraph{Bidirectional Typechecking} The methods \lstinlinep{ctx.ana} and \lstinlinep{ctx.syn} are also defined by the compiler and can be called by type constructors to request type analysis (when the expected type is known) and synthesis (when the type is an ``output''), respectively, for an expression.
+\paragraph{Expressions} The methods \lstinlinep{ctx.ana} and \lstinlinep{ctx.syn} are also defined by the compiler and can be called by type constructors to request type analysis (when the expected type is known) and synthesis (when the type is an ``output''), respectively, for an expression.
 Once again, the compiler delegates to a type constructor by a protocol that depends on the term constructor, invoking methods of the form \lstinlinep{ana_TermCon} or \lstinlinep{syn_TermCon}.
@@
 
-\paragraph{Names} If the term constructor of an expression is \lstinlinep{ast.Name}, as when referring to a bound variable or an assignable location, then the type constructor governing the function the term appears in is delegated to via the class method \lstinlinep{syn_Name} (names, when used as expressions, must synthesize a type).
+\subparagraph{Names} If the term constructor of an expression is \lstinlinep{ast.Name}, as when referring to a bound variable or an assignable location, then the type constructor governing the function the term appears in is delegated to via the class method \lstinlinep{syn_Name} (names, when used as expressions, must synthesize a type).
 We see the definition of this method for \lstinlinep{fn} starting on line \ref{syn_Name} in Listing \ref{fn-stmts}.
@@
 In the former case, the type that was synthesized when the assignment occurred (by \lstinlinep{syn_Assign_Name}) is returned.
-In the latter case, we must lift the static value to run-time.
-For the purposes of this paper, we support only other typed @\texttt{lang} functions, Python functions and classes (which are given type \lstinlinep{dyn}, and can only be called with values of type \lstinlinep{dyn}) and modules (which are given a \emph{singleton type} -- a type indexed by the module reference itself -- from which @\text{lang} functions and Python functions and classes can be accessed as attributes in the usual way, see below).%In the future, we plan to detail how the interface between Python and @\texttt{lang} can be enriched to support moving typed values in and out of untyped Python functions safely (by inserting checks, defined again by a type constructor, at the interface between typed and untyped code), but we do not have enough room to adequately detail this mechanism here (and it is tangential to our specific aims).
-\paragraph{Literal Expressions and Ascriptions}
+
+In the latter case, we must lift the static value to run-time and give it an appropriate type.
+This cannot always be done automatically (because it is a form of serialization).
+Instead, any Python value with a \lstinlinep{atlang_syn_Lift} method returning an \lstinlinep{atlang} type can provide explicit support for this operation.
+A few other constructs also have built in support.
+For the purposes of this section, this includes other typed @\texttt{lang} functions (which synthesize their own type), untyped Python functions and classes (which synthesize types \lstinlinep{pyfun} and \lstinlinep{pyclass}, and thus can only be called with values of type \lstinlinep{dyn}), modules (which are given a \emph{singleton type} -- a type indexed by the module reference itself -- from which other values that can be lifted can be accessed as attributes, see below), and immutable Python data structures like numbers, strings and tuples.
+%In the future, we plan to detail how the interface between Python and @\texttt{lang} can be enriched to support moving typed values in and out of untyped Python functions safely (by inserting checks, defined again by a type constructor, at the interface between typed and untyped code), but we do not have enough room to adequately detail this mechanism here (and it is tangential to our specific aims).
+\subparagraph{Literal Expressions and Ascriptions}
 \begin{codelisting}[t]
@@
 The various literal forms inside the outermost form thus do not need an ascription because they appear in positions where the type they will be analyzed against is provided by \lstinlinep{record}.
-For example, the value of the field \lstinlinep{account_num} delegates to \lstinlinep{string_in} via the \lstinlinep{ana_Str} method (not shown).
+For example, the value of the field \lstinlinep{account_num} delegates to \lstinlinep{string_in} via the \lstinlinep{ana_Str} method, shown in Listing \ref{string_in}.
 
@@
 A different function type constructor might make more precise choices.
-Indeed, a benefit of using Python as our static language is that it is relatively straightforward to provide a type constructor that allows us to provide different defaults, and control choices like the return semantics, as block-scoped ''pragmas'' in the static language using Python's \lstinlinep{with} statement \cite{python} (details omitted), e.g.
+Indeed, a benefit of using Python as our static language is that it is relatively straightforward to define a type constructor that allows for different defaults using block-scoped ''pragmas'' in the static language via Python's \lstinlinep{with} statement \cite{python} (details omitted), e.g.
 \begin{codelisting}[h]
@@
 
-\paragraph{Targeted Expressions} Expression forms having exactly one sub-expression, like \lstinlinep{-e} (term constructor \lstinlinep{UnaryOp_USub}) or \lstinlinep{e.attr} (term constructor \lstinlinep{Attribute}), or where there may be multiple sub-expressions but the left-most one is the only one required, like \lstinlinep{e[slices]} (term constructor  \lstinlinep{Subscript}) or \lstinlinep{e(args)} (term constructor \lstinlinep{Call}), are called \emph{targeted expressions}.
+\subparagraph{Targeted Expressions} Expression forms having exactly one sub-expression, like \lstinlinep{-e} (term constructor \lstinlinep{UnaryOp_USub}) or \lstinlinep{e.attr} (term constructor \lstinlinep{Attribute}), or where there may be multiple sub-expressions but the left-most one is the only one required, like \lstinlinep{e[slices]} (term constructor  \lstinlinep{Subscript}) or \lstinlinep{e(args)} (term constructor \lstinlinep{Call}), are called \emph{targeted expressions}.
 For these, the compiler first synthesizes a type for the left-most sub-expression, then calls either the \lstinlinep{ana_TermCon} or \lstinlinep{syn_TermCon} methods on that type.
@@
 
-\paragraph{Binary Expressions} \begin{codelisting}[t]
+\subparagraph{Binary Expressions} \begin{codelisting}[t]
 \begin{lstlisting}
@@
     atlang.Type.__init__(self, self._rlang_of_rx(idx))
+
+  def ana_Str(self, ctx, e):
+    _check_rx(self.idx, e.s)
+
+  @classmethod
+  def syn_idx_Str(cls, ctx, e):
+    return _rx_of_str(e.s) # most specific rx
 
@@
 This is a very simple modular analysis (rather than one that can only be performed at ``link-time''), shown in Listing \ref{check_tycon}, that ensures that the type constructor delegated control over each binary expression is deterministic and  unambiguous without arbitrarily preferring one subexpression position over another.
-This check is performed  by using a metaclass hook (technically, this can be disabled; we assume that clients are not importing potentially adversarial extensions in this work, though lifting this assumption raises quite interesting questions that we hope will be addressed by future work).
+This check is performed  by using a metaclass hook.% (technically, this can be disabled; we assume that clients are not importing potentially adversarial extensions in this work, though lifting this assumption raises quite interesting questions that we hope will be addressed by future work).
 
@@
 Examples of translation methods for the \lstinlinep{fn} and \lstinlinep{record} type constructors are shown in Listing \ref{trans}.
-The output of translation on our example is discussed in the next subsection.
+The output of translation for our example is discussed in the next subsection.
 Translation methods have access to the context and node, as during typechecking, and must return a translation, which for our purposes, is simply another Python syntax tree (in practice, we offer some additional conveniences beyond \lstinlinep{ast}, such as fresh variable generation and lifting of imports and statements inside expressions to appropriate positions, in the module \lstinlinep{astx} distributed with the language).
@@
     if e.id in ctx.assn_ctx: return astx.copy(e)
-    else: return self._trans_lift(
+    else: return ctx.trans_Lift(
       ctx.static_env[e.id])
@@
     ast_dict = dict(zip(e.keys, e.values))
-    return target.Tuple(ctx.trans(target, ast_dict[f])
+    return ast.Tuple((f, ctx.trans(ast_dict[f]))
       for f, ty in self.idx)
@@
 \item Terms of type \lstinlinep{string_in[r"..."]} translated to strings.
-Checks could here be performed  statically.% Note that there are situations (e.g.
+Checks were all performed  statically in this example.% Note that there are situations (e.g.
 if a string is read in from the console) that would necessitate an initial run-time check, but no further checks in downstream functions.
@@
 In @\texttt{lang}, these can be found without executing the code (i.e.
-true static typechecking).
+we have true static typechecking).
 
@@
 
-In future work, we plan to detail how to insert dynamic checks and wrapper objects, defined by active type constructors in a manner similar to how static checks are defined here, so that values that are not of type \lstinlinep{dyn} can be passed in and out of untyped Python code.
+If a function captures any static values that do not support lifting, then that function can only be used via interactive invocation (there is no way to separate compile-time from run-time without lifting captured values).
+Such values when captured in an interactively invoked setting by default synthesize type \lstinlinep{dyn}, though they can provide a \lstinline{atlang_syn_Capture} method that synthesizes a more specific type.
+We will see an example of this and a related use of phaseless capture in our discussion of \lstinlinep{atlib.opencl} below.
+
+%\todo{move to future work section?}In future work, we plan to detail how to insert dynamic checks and wrapper objects, defined by active type constructors in a manner similar to how static checks are defined here, so that values that are not of type \lstinlinep{dyn} can be passed in and out of untyped Python code.
 Because these can be seen as implicit coercions to and from \lstinlinep{dyn}, and we do not aim to introduce this feature in this paper, we omit discussion of this topic.
@@
 \section{More Examples}\label{examples}
-In the previous section, we showed examples of several interesting type system fragments implemented as libraries using \verb|atlang|, including functional record types, a prototypic object system and regular string types.
-A more detailed description of regular string types and their implementation using \verb|atlang| was recently published \cite{sanitation-psp14}.
-Here, we showcase two more powerful examples that demonstrate the expressive power of system: functional datatypes with nested pattern matching, and a type safe foreign function interface to the complete OpenCL language for many-core programming on devices like GPUs.
+In the previous section, we showed examples of several interesting type system fragments implemented as libraries using \lstinlinep{atlang}, including functional record types, a prototypic object system and regular string types.
+A more detailed description of regular string types and their implementation using \lstinlinep{atlang} was recently published at a workshop \cite{sanitation-psp14}.
+Here, we showcase two more powerful examples that demonstrate the expressive power of the system: functional datatypes with nested pattern matching, and a type safe foreign function interface to the complete OpenCL language for many-core programming on devices like GPUs.
 
 \subsection{Functional Datatypes and Nested Pattern Matching}
-A powerful feature of modern functional languages like ML is support for nested pattern matching over datatypes (i.e.
-recursive labeled disjoint sums) and tuples.
-For example, tree structures are well-modeled in this way, as we show in Listing \ref{example-casetypes}.
-Here, \verb|Tree| is a \emph{case type}, which is what we call functional datatypes in \verb|atlib|.
-The type constructor \verb|Tree| is declared as a named type using the \verb|@ty| annotation.
-This is a general mechanism in \verb|atlang| for supporting recursive types that behave generatively, i.e.
-that are identified by a name, rather than a structure.
-Functional datatypes have the same character.
-The \verb|@ty| annotation also supports parameterized types.
-Here, \verb|a| must be another type.
-The type index is a series of cases with, optionally, corresponding types.
-Here, a (binary) tree can either be empty (no associated data), a leaf with an associated value of type \verb|a|, or an internal node, which takes a pair of trees as data.
-We use the \verb|tuple| type constructor to represent pairs.
-
-The function \verb|treesum| on lines 11-17 takes a tree containing \verb|dyn| values and computes the sum of these values.
-The case operator is defined by the \verb|casetype| type constructor, leveraging a combination of the \verb|Attribute| and the \verb|Subscript|  term constructors (both are targeted operations, see above).
-The slice syntax is again re-interpretted, here s a series of rules, each of which starts with a pattern, followed by a colon, then a case.
-Variables in patterns can be bound in the case.
-Patterns can consist of case names and tuple patterns at arbitrary depth.
-More generally, we have done initial designs for an extensible pattern language, again delegating to type constructors to determine whether the pattern is valid and which variables it binds, but here we restrict ourselves to these two type constructors.
-
-By combining the functional behavior of the \verb|fn| type constructor, which dispenses with ``return'' statements, and creatively repurposing existing syntax, we are thus able to implement an essentially idiomatic statically-typed functional language, entirely as a library for Python.
-The constructs we defined can be composed arbitrarily with, e.g.
-the record types or regular string types in the previous section, without any concern regarding syntactic or semantic conflict, because \verb|casetype| only controls the semantics of the \verb|case| operation on values that are of such a type.
-
-\subsection{A Low-Level Foreign Function Interface to OpenCL} 
 \begin{codelisting}[t]
@@
 from atlang import ty
-from atlib import casetype, tuple, fn
+from atlib import case, casetype, tuple, fn
 
@@
 
-@fn
+@fn # using Python 3 type annotation syntax
 def treesum(tree : Tree(dyn)) -> dyn:
-  tree.case[
+  case(tree) [
     Empty: 0,
@@
   ]
+
+@fn
+def __main__():
+  treesum(Tree.Node(Tree.Leaf(5), Tree.Leaf(5)))
 \end{lstlisting}
@@
 \end{codelisting}
+\begin{codelisting}[t]
+\begin{lstlisting}
+def treesum(tree):
+  return (0 if tree[0] == 0 else 
+    tree[1] if tree[0] == 1 else 
+    treesum(tree[1][0]) + treesum(tree[1][1]))
+
+def __main__():
+  treesum((2, ((1, 5), (1, 5))))
+\end{lstlisting}
+\caption{The translation of Listing \ref{example-casetypes}. Case types are implemented as fast tagged values.}
+\label{example-casetypes-out}
+\end{codelisting}
+\begin{codelisting}[t]
+\begin{lstlisting}
+class _case(object):
+  def syn_Lift(self):
+    return caseop[()]
+case = _case()
+
+class caseop(atlang.Type):
+  def syn_Call(self, ctx, e):
+    ctx.syn(e.args[0])
+    if len(e.args) != 0:
+      raise atlang.TypeError("...", e)
+    return caseop_applied[()]
+
+class caseop_applied(atlang.Type):
+  def syn_Subscript(self, ctx, e):
+    scrutinee = e.value.args[0]
+    rules = _rules_from(e.slices)
+    ty = None
+    for rule in rules:
+      bindings = bindings_from_pat(rule, scrutinee.ty)
+      _push_bindings(ctx, bindings)
+      cur_type = ctx.syn(rule.branch)
+      _pop_bindings(ctx, bindings)
+      if ty is not None and cur_type != ty:
+        raise atlang.TypeError("inconsistent", rule)
+      else: ty = cur_type
+    return ty
+
+  def trans_Subscript(self, ctx, e):
+    scrutinee = e.value.args[0]
+    return scrutinee.ty.trans_pats(
+      _rules_from(e.slices))
+
+class casetype(atlang.Type):
+  # ...
+  def type_pat_Call(self, ctx, p):
+    if p.func in self.idx: 
+      return self.idx[p.func]
+
+  def trans_pats(self, ctx, pats):
+    # ... turns sequence of patterns at same level 
+    # into a conditional expr...
+
+class tuple(atlang.Type):
+  # ...
+  def type_pat_Tuple(self, ctx, p):
+    if len(p.elts) == len(self.idx):
+      return self.idx
+    else raise PatError("invalid pat length") 
+\end{lstlisting}
+\caption{The implementation of the case analysis operator uses intermediate type constructors that contain only typing logic but no translation logic.
+It also defines its own ``second-order'' extensibility mechanism.}
+\label{example-casetypes-impl}
+\end{codelisting}
+A powerful feature of modern functional languages like ML is support for nested pattern matching over datatypes (i.e.
+recursive labeled sum types), tuples, records and other types.
+For example, tree structures are well-modeled using recursive labeled sums, as we show in Listing \ref{example-casetypes}.
+Here, \lstinlinep{Tree} is a \emph{case type}, which is what we call labeled sums in \lstinlinep{atlib}.
+\lstinlinep{Tree} is declared as a \emph{named type} using the \lstinlinep{@ty} annotation.
+This is a general mechanism in \lstinlinep{atlang} for supporting recursive types that behave generatively, i.e.
+that are identified by a name.
+Functional datatypes in ML and similar languages are also generative in this way.
+The \lstinlinep{@ty} mechanism also supports parameterized types: here, \lstinlinep{a} stands in for any other type.
+The type index to \lstinlinep{casetype} is a series of cases with, optionally, associated types.
+Here, a (binary) tree can either be empty (no associated data), a leaf with an associated value of type \lstinlinep{a}, or an internal node, which takes a pair of trees as data.
+We use the \lstinlinep{atlib.tuple} type constructor to represent pairs.
+
+The function \lstinlinep{treesum} on lines 11-17 takes a tree containing \lstinlinep{dyn} values and computes the sum of these values by nested pattern matching.
+The \lstinlinep{__main__} function constructs a tree and calls \lstinlinep{treesum} on it.
+The result of compiling Listing \ref{example-casetypes} is shown in Listing \ref{example-casetypes-out}.
+Case types translate to pairs consisting of a numeric tag and data, and case analysis translates to nested conditionals.
+
+Some relevant portions of the implementation of case analysis are shown in Listing \ref{example-casetypes-impl}.
+The object \lstinlinep{case} is an instance of the class \lstinlinep{_case}, which defines the method \lstinlinep{syn_Lift}.
+When the imported name \lstinlinep{case} is encountered in Listing \ref{example-casetypes}, this is called, per the discussion earlier on names in the static environment.
+The result is that \lstinline{case} synthesizes type \lstinlinep{caseop[()]}.
+
+The only purpose of the type constructor \lstinlinep{caseop} is to define the \lstinlinep{syn_Call} method, which ensures that the case operator is applied to a single well-typed scrutinee.
+The combined term \lstinlinep{case(tree)} then synthesizes type \lstinlinep{caseop_applied[()]}.
+Notice that neither method is paired with a \lstinlinep{trans_} method -- these types are merely ``bookkeeping'' for the compound case analysis form.
+Parts of this form are not meaningful by themselves, so translation would fail if, for example, \lstinlinep{case(tree)} was written by itself without cases.
+
+The \lstinlinep{caseop_applied} tycon defines the \lstinlinep{syn_Subscript} method, which actually processes the list of provided rules.
+First, the type of the scrutinee is asked to process the bindings from the pattern portion of each rule.
+This involves calling methods of the form \lstinlinep{type_pat_TermCon}, which must return types for all the sub-patterns.
+For example, the \lstinline{type_pat_Call} method of \lstinline{casetype} makes sure that the case named is defined, and if so returns the associated type, taken from its type index, as the type of the inner pattern.
+The process repeats recursively, so that a tuple pattern might then be processed by \lstinlinep{type_pat_Tuple}.
+Any type constructor can participate in this protocol by defining appropriate methods, so it is extensible.
+Note, however, that this is a ``second-order'' extensibility mechanism -- nothing in \verb|atlang| itself supports pattern matching.
+Instead, \lstinlinep{case} itself defined this protocol in a library, demonstrating the power of using a flexible general-purpose language for writing extensions.
+
+The translation phase begins at the \lstinlinep{trans_Subscript} method of \lstinlinep{caseop_applied}.
+This simply hands responsibility for translating the entire set of rules to the type of the scrutinee.
+This may then hand control to the types of inner patterns (analagous to the typing phase, not shown).
+
+By combining the functional behavior of the \lstinlinep{fn} type constructor, which dispenses with ``return'' statements, and creatively repurposing existing syntax, we are thus able to implement an essentially idiomatic statically-typed functional language, entirely as a library for Python.
+The constructs we defined can be composed arbitrarily with, e.g.
+the record types, object types or regular string types in the previous section, or the typed OpenCL FFI we will describe next, without any concern regarding syntactic or semantic conflicts.
+This is enabled entirely by taking a type constructor oriented view toward the extensibility problem.
+
+\subsection{A Low-Level Foreign Function Interface to OpenCL}
 
@@
 import atlib.opencl as opencl
-
-dev = opencl.init_device(0)
-buffer = dev.send(numpy.zeros((80000,)))
-buffer_t = buffer.ty # == opencl.global_ptr[elem_t]
-elem_t = buffer_t.elem_t # == opencl.double
-
-@opencl.fn
-def map(x : buffer_t, f : opencl.fn[elem_t, elem_t]):
-  gid = get_global_id(0)
-  x[gid] = f(x[gid])
-
-@opencl.fn
-def mad(x : elem_t, m : elem_t, y : elem_t):
-  return m*x + y
-@opencl.fn
-def my_mad(x : elem_t): return mad(x, 10, 5)
-@opencl.fn
-def my_mad2(x : elem_t): return mad(x, 20, 5)
-
-map(buffer, my_mad)
-map(buffer, my_mad2)
+import numpy
+
+device = opencl.init_device(0)
+buffer = device.send(numpy.zeros((800000,))
+
+@opencl.template_fn
+def map(b, f):
+  gid = get_global_id(0) # data parallel thread ID
+  b[gid] = f(b[gid])
+
+@opencl.template_fn
+def shift(x):
+  return x + 5
+
+@opencl.template_fn
+def triple(x):
+  return 3 * x
+
+map(buffer, triple)
+map(buffer, shift)
 \end{lstlisting}
-\caption{An example of a type system for OpenCL.}
+\caption{An example use of our typed FFI to OpenCL, demonstrating both template functions and phaseless capture.}
 \label{example-opencl}
@@
 \begin{codelisting}[t]
-\begin{lstlisting}
+\begin{lstlisting}[style=OpenCL]
 #pragma OPENCL EXTENSION cl_khr_fp64 : enable
 
-double mad(double x, double m, double y) {
-  return m*x + y;
+double shift(double x) {
+  return x + 5.0;
 }
 
-double my_mad(double x) {
-  return mad(x, 10.0, 5.0);
+double triple(double x) {
+  return 3.0 * x;
 }
 
-double my_mad2(double x) {
-  return mad(x, 20.0, 5.0);
+kernel void map__0_(global double* b) {
+  size_t gid = get_global_id(0); 
+  b[gid] = triple(b[gid])
 }
 
-kernel void map__1_(double x) {
+kernel void map__1_(global double* b) {
   size_t gid = get_global_id(0);
-  x[gid] = my_mad(x[gid]);
-}
-
-kernel void map__2_(double x) {
-  size_t gid = get_global_id(0);
-  x[gid] = my_mad2(x[gid]);
+  b[gid] = shift(b[gid])
 }
 \end{lstlisting}
-\caption{The underlying code generated by atlib.opencl.}
+\caption{The underlying code generated by \texttt{atlib.opencl} as a string passed through \texttt{pyopencl}.}
 \label{example-opencl-out}
 \end{codelisting}
+\begin{codelisting}[t]
+\begin{lstlisting}
+class template_fn(atlang.Type):
+  # ...
+  def syn_Call(self, ctx, e):
+    arg_types = [ctx.syn(arg) for arg in e.args]
+    return fn[fn.syn_idx_FunctionDef(ctx, self.idx.fn, 
+      (arg_types, Ellipsis))]
+
+  def trans_Call(self, ctx, e):
+    return _wrap_in_pyopencl(
+      e.ty.trans_FunctionDef(ctx, self.idx.fn))
+\end{lstlisting}
+\caption{A portion of the logic of OpenCL template functions, showing how they defer to the logic for standard OpenCL functions at each call site, rather than at the declaration site.}
+\label{example-opencl-templatefn}
+\end{codelisting}
+\begin{codelisting}[t]
+\begin{lstlisting}
+class Buffer(pyopencl.Buffer):
+  # ...
+  def atlang_syn_Capture(self):
+    return global_ptr[
+      _np_dtypes_to_cltypes(self.dtype)]
+\end{lstlisting}
+\caption{The \texttt{opencl.Buffer} class represents OpenCL memory objects in global device memory, inheriting from the \texttt{pyopencl.Buffer} class.
+These supports phaseless capture at the corresponding global pointer type.}
+\label{example-opencl-buffer}
+\end{codelisting}
 
 Python is a common language in scientific computing and data analysis domains.
-The performance of large-scale analyses can often be a bottleneck, so programmers often wish to write and interact with code written in a low-level language.
-OpenCL is designed precisely for this workflow, exposing a C-based language that can compile to a variety of specialized hardware (e.g.
-GPUs) via a standard API.
-The \verb|pyopencl| package exposes this API to Python code and integrates it with the popular \verb|numpy|numeric array package \cite{klockner2011pycuda}.
+The performance of large-scale analyses can be a bottleneck, so programmers often wish to write and interact with code written in a low-level language.
+OpenCL is designed precisely for this workflow, exposing a C-based data parallel language that can compile to a variety of specialized hardware (e.g.
+GPUs) via a standard API, as well as a runtime that allows a program to manage device memory.
+The \lstinlinep{pyopencl} package exposes these APIs to Python code and integrates them with the popular \lstinlinep{numpy} numeric array package \cite{klockner2011pycuda}.
 To compile an OpenCL function, however, users must write it as a string.
 This is neither idiomatic nor safe, because it defers typechecking to invocation-time.
-
-In Listing \ref{example-opencl}, we show usage of the \verb|atlib.opencl| package, which implements the entirety of the OpenCL language (which includes essentially the entirety of C99, plus some extensions to work with multiple memory spaces).
-Although the details are beyond the scope of this paper, we note that the mapping onto Python syntax was straightforward, particularly given the flexibility of analytic numeric literal forms, as described above.
-
-After initializing a device, line 4 sends a \verb|numpy| array to the device, assigning \verb|buffer| its handle.
-Line 5 extracts its type, which is based on the dynamic element type of the \verb|numpy| array.
-The type constructor \verb|global_ptr| defines the semantics of OpenCL's pointers to global memory (e.g.
-subscripting, pointer arithmetic, pointer differencing), consistent with the C99 specification.
-Because the compilation script is programmatic, we can extract its element type programmatically as well.
-Here, it is the \verb|opencl.double| type, which supports arithmetic operations, including the full array of widening rules defined by C99 when paired with any of the other numeric OpenCL types (using handle sets, above, to prioritize control to the wider type).
-
-Lines 8-19 define several OpenCL functions using Python syntax, but retaining OpenCL's semantics.
-The \verb|map| function takes a buffer argument and a function over its elements and applies the function to each element in the buffer in a data parallel manner.
-Each thread handles the element corresponding to its thread ID, called a \verb|group ID| in OpenCL and accessed by a primitive on the first line.
-Note that OpenCL does not itself have support for higher-order functions, however.
-Instead, higher order functions are essentially an extension to the foreign language being encoded using active type constructors.
-The OpenCL function decorator treats uses of higher order functions as if each function ever provided to it had a \emph{singleton type}, i.e.
-a type inhabited only by that particular function.
-Each call lazily generates code for a specialized variant of that function.
-The result can be seen in Listing \ref{example-opencl-out}, which shows the OpenCL code generated (as a string sent through \verb|pyopencl|) by this example.
+OpenCL is also often too low-level, not supporting even features found in other low-level languages like function overloading, function templates and higher-order functions.
+
+In Listing \ref{example-opencl}, we show usage of the \lstinlinep{atlib.opencl} package, which implements the entirety of the OpenCL type system (which includes essentially the entirety of C99, plus some extensions to work with multiple memory spaces).
+Although the details are beyond the scope of this paper, we note that the mapping onto Python syntax was straightforward, particularly given the flexibility of analytic numeric literal forms, as described above, to support the wide variety of number types in C99-based languages.
+
+After initializing a device, line 4 sends a \lstinlinep{numpy} array to the device, assigning \lstinlinep{buffer} its handle, an object inheriting from \lstinlinep{opencl.Buffer}.
+The type of the \lstinlinep{numpy} array is also extracted and stored in this object.
+
+On lines 7-18, we define three OpenCL template function.
+Let us consider the first, \lstinlinep{map}, which takes two arguments.
+A template function is one that does not have specified argument types, generally because many types might be valid, as in this case.
+The body of \lstinlinep{map} would be valid for any combination of types for \lstinlinep{b} and \lstinlinep{f} that support subscripting and calls as shown.
+Template functions thus defer typechecking to every invocation site, based on the types synthesized by the provided arguments.
+This is analagous to template functions in C++, but here we implicitly assume a distinct template parameter for each argument and implicitly instantiate these parameters at use sites.
+Technically, such functions have a singleton type, indexed by the function definition itself, as shown in Listing \ref{example-opencl-templatefn}.
+Note that  OpenCL itself doesn't have template functions; we are implementing them, essentially by extending the language across its FFI.
+
+We see two applications of \lstinlinep{map} on lines 20 and 21.
+Note that these are in the dynamically typed portion of the code.
+By default, the first argument in each case, \lstinlinep{buffer}, would synthesize type \lstinlinep{dyn} and typechecking would fail (at line 10, technically, because \lstinlinep{dyn} doesn't support subscripting with the type that \lstinlinep{gid} synthesizes).
+But we make use of the support for phaseless capture discussed in Sec.
+\ref{interactive}, which allows values to define a method, \lstinlinep{atlang_syn_Capture}, that permits capture and, in this case, function application with values that would not otherwise support crossing into \lstinlinep{atlang} functions with argument types other than \lstinlinep{dyn}.
+In this case, \lstinlinep{Buffer} objects synthesize a corresponding pointer type, represented by the tycon \lstinlinep{opencl.global_ptr} and indexed by the target type of the pointer, here \lstinlinep{opencl.double} because the initial \lstinlinep{numpy} array was an array of \lstinlinep{double}s and there is a mapping from \lstinlinep{numpy} data types to OpenCL types.
+The implementation is shown in Listing \ref{example-opencl-buffer}.
+Note that phaseless capture is distinct from lifting -- lifting requires a value persist from compile-time to run-time, but OpenCL buffers are clearly ephemeral.
+Phaseless capture can only occur when interactive invocation is being used.
+
+Each invocation of \lstinlinep{map} received the same buffer, but a different function.
+Because these were template functions, their types were distinct.
+As such, two different versions of map were generated, as shown in Listing \ref{example-opencl-out}.
+Indeed, without support for function pointers in OpenCL, code generation like this is the standard way of supporting ``higher-order'' functions like \lstinlinep{map}.
+Rather than having to do this manually, however, our type system handled this internally.
 
@@
 In many cases, the typechecking code can be inherited directly.
-This is, we argue, rather compelling: a decidedly low-level language like OpenCL can be extended with low-overhead versions of sophisticated features, like a prototypic object system, modularly, via its foreign function interface from a scripting language like Python.
+This is, we argue, rather compelling: a decidedly low-level language like OpenCL can be extended with low-overhead versions of sophisticated features, like a prototypic object system or case types, modularly, via its foreign function interface from a scripting language like Python.
 
@@
 Static terms have a \emph{static dynamics} and evaluate to \emph{static values} or fail.
+Static terms are analagous to top-level Python code in \verb|atlang|, and external terms are analagous to terms inside typed functions.
 
@@
 
-Recent work we have done on the Wyvern programming language uses bidirectional typechecking to control aspects of literal syntax in a manner similar to, but somewhat more flexible than, how we treat introductory forms \cite{TSLs}.
+In recent work on the Wyvern programming language, we also used bidirectional typechecking to control aspects of literal syntax in a manner similar to, but somewhat more syntactically flexible than, how we treat introductory forms \cite{TSLs}.
 Wyvern is its own language dialect and does not have an extensible type system, supporting only desugarings like SugarJ \cite{erdweg2011sugarj}.
@@
 security-oriented types) using our framework.
+Type systems that require tracking contextual information are possible using our framework, but the context needs to be initialized on first use of a type that needs it.
+This is somewhat awkward, so we plan to include a mechanism that allows any imported tycons to initialize the context ahead of typechecking.
 
@@
 An even more interesting goal would be to guarantee that extensions are mutually conservative: that one cannot weaken any of the guarantees of the other.
-We believe that by enforcing strict abstraction barriers between extensions, we can approach this goal.
-Bootstrapping the @\texttt{lang} compiler would be an interesting direction to explore to enable this.
+We believe that by enforcing strict abstraction barriers between extensions, we can approach this goal (and a manuscript is in preparation).
+Bootstrapping the @\texttt{lang} compiler would be an interesting direction to explore, to avoid the awkwardness of relying on a loose dynamically typed language to implement static type systems.
 % Different dialects can be difficult to combine manually, as we discussed in the introduction.
@@
 % We recommend abbrvnat bibliography style.
-
+\newpage
 \bibliographystyle{abbrv}

### 2014-10-16 sanitation-psp14/tr/strings_chapter.tex
Added TR.

\documentclass[12pt]{article}

\usepackage{lineno}

\usepackage{afterpage}
\usepackage{listings}
\usepackage[usenames,dvipsnames]{color}
\usepackage{mathpartir}
\usepackage{amsmath}
\usepackage{amssymb}

% Hack. Something's wrong with PLAS paper when using the ACM Proc docclass
\let\proof\relax
\let\endproof\relax
\usepackage{amsthm}

\usepackage[nomarkers,figuresonly]{endfloat}
\renewcommand{\efloatseparator}{\vfill}

\usepackage{ stmaryrd }
\usepackage{verbatimbox}
\input{../att-icfp14/macros-atlam}
\usepackage{alltt}
\renewcommand{\ttdefault}{txtt}
\usepackage{color}
\usepackage[usenames,dvipsnames,svgnames,table]{xcolor}
\definecolor{mauve}{rgb}{0.58,0,0.82}
\definecolor{light-gray}{gray}{0.5}
\usepackage{listings}
\usepackage{wasysym}
    \makeatletter
\usepackage{enumitem}
\usepackage{enumerate}

% \btIfInRange{number}{range list}{TRUE}{FALSE}
%
% Test if int number <number> is element of a (comma separated) list of ranges
% (such as: {1,3-5,7,10-12,14}) and processes <TRUE> or <FALSE> respectively
%
        \newcount\bt@rangea
        \newcount\bt@rangeb

        \newcommand\btIfInRange[2]{%
            \global\let\bt@inrange\@secondoftwo%
            \edef\bt@rangelist{#2}%
            \foreach \range in \bt@rangelist {%
                \afterassignment\bt@getrangeb%
                \bt@rangea=0\range\relax%
                \pgfmathtruncatemacro\result{ ( #1 >= \bt@rangea) && (#1 <= \bt@rangeb) }%
                \ifnum\result=1\relax%
                    \breakforeach%
                    \global\let\bt@inrange\@firstoftwo%
                \fi%
            }%
            \bt@inrange%
        }

        \newcommand\bt@getrangeb{%
            \@ifnextchar\relax%
            {\bt@rangeb=\bt@rangea}%
            {\@getrangeb}%
        }

        \def\@getrangeb-#1\relax{%
            \ifx\relax#1\relax%
                \bt@rangeb=100000%   \maxdimen is too large for pgfmath
            \else%
                \bt@rangeb=#1\relax%
            \fi%
        }

%
% \btLstHL{range list}
%
        \newcommand{\btLstHL}[1]{%
            \btIfInRange{\value{lstnumber}}{#1}%
            {\color{black!10}}%
            {\def\lst@linebgrd}%
        }%

%
% \btInputEmph[listing options]{range list}{file name}
%
        \newcommand{\btLstInputEmph}[3][\empty]{%
            \lstset{%
                linebackgroundcolor=\btLstHL{#2}%
                \lstinputlisting{#3}%
            }% \only
        }

% Patch line number key to call line background macro
        \lst@Key{numbers}{none}{%
            \def\lst@PlaceNumber{\lst@linebgrd}%
            \lstKV@SwitchCases{#1}{%
                none&\\%
                left&\def\lst@PlaceNumber{\llap{\normalfont
                \lst@numberstyle{\thelstnumber}\kern\lst@numbersep}\lst@linebgrd}\\%
                right&\def\lst@PlaceNumber{\rlap{\normalfont
                \kern\linewidth \kern\lst@numbersep
                \lst@numberstyle{\thelstnumber}}\lst@linebgrd}%
            }{%
                \PackageError{Listings}{Numbers #1 unknown}\@ehc%
            }%
        }

% New keys
        \lst@Key{linebackgroundcolor}{}{%
            \def\lst@linebgrdcolor{#1}%
        }
        \lst@Key{linebackgroundsep}{0pt}{%
            \def\lst@linebgrdsep{#1}%
        }
        \lst@Key{linebackgroundwidth}{\linewidth}{%
            \def\lst@linebgrdwidth{#1}%
        }
        \lst@Key{linebackgroundheight}{\ht\strutbox}{%
            \def\lst@linebgrdheight{#1}%
        }
        \lst@Key{linebackgrounddepth}{\dp\strutbox}{%
            \def\lst@linebgrddepth{#1}%
        }
        \lst@Key{linebackgroundcmd}{\color@block}{%
            \def\lst@linebgrdcmd{#1}%
        }

% Line Background macro
        \newcommand{\lst@linebgrd}{%
            \ifx\lst@linebgrdcolor\empty\else
                \rlap{%
                    \lst@basicstyle
                    \color{-.}% By default use the opposite (`-`) of the current color (`.`) as background
                    \lst@linebgrdcolor{%
                        \kern-\dimexpr\lst@linebgrdsep\relax%
                        \lst@linebgrdcmd{\lst@linebgrdwidth}{\lst@linebgrdheight}{\lst@linebgrddepth}%
                    }%
                }%
            \fi
        }

 % Heather-added packages for the fancy table
 \usepackage{longtable}
 \usepackage{booktabs}
 \usepackage{pdflscape}
 \usepackage{colortbl}%
 \newcommand{\myrowcolour}{\rowcolor[gray]{0.925}}
 \usepackage{wasysym}

    \makeatother

\lstset{
  language=Python,
  showstringspaces=false,
  formfeed=\newpage,
  tabsize=4,
  commentstyle=\itshape\color{light-gray},
  basicstyle=\ttfamily\scriptsize,
  morekeywords={lambda, self, assert, as, cls},
  numbers=left,
  numberstyle=\scriptsize\color{light-gray}\textsf,
  xleftmargin=2em,
  stringstyle=\color{mauve}
}
\lstdefinestyle{Bash}{
    language={}, 
    numbers=left,
    numberstyle=\scriptsize\color{light-gray}\textsf,
    moredelim=**[is][\color{blue}\bf\ttfamily]{`}{`},
}
\lstdefinestyle{OpenCL}{
	language=C++,
	morekeywords={kernel, __kernel, global, __global, size_t, get_global_id, sin, printf, int2}
}

\usepackage{float}
\floatstyle{ruled}
\newfloat{codelisting}{tp}{lop}
\floatname{codelisting}{Listing}
\setlength{\floatsep}{10pt}
\setlength{\textfloatsep}{10pt}

\usepackage{url}

%\usepackage{todo}
%\usepackage[subject={Todo},author={Josef}]{pdfcomment}
%\usepackage{cooltooltips}
\newcommand{\todo}[1]{{\color{red} #1}}
\usepackage{placeins}

\usepackage{textpos}

\renewcommand\topfraction{0.85}
\renewcommand\bottomfraction{0.85}
\renewcommand\textfraction{0.1}
\renewcommand\floatpagefraction{0.85}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\usepackage{bussproofs}    % Gentzen-style deduction trees *with aligned sequents!*
\usepackage{mathpartir}	% For type-settting type checking rule figures
\usepackage{syntax}		% For type-setting formal grammars.

\usepackage{hyperref}		% For links in citations

\usepackage{float}			% Make figures float if [H] option is passed.

\iffalse
\usepackage{listings}		% For typesetting code listings
\usepackage{callout}
\usepackage{titlesec}
\usepackage[T1]{fontenc}
\usepackage{upquote}
\lstset{upquote=true}
\fi

\usepackage{textcomp}		% For \textquotesingle as used in introduction
\usepackage{color}			% for box colors, like in TAPL.

\usepackage{amsmath}		% Begin Carthage default packages
\usepackage{makeidx}
\usepackage{amsthm}
\usepackage{amsfonts}
\usepackage{amssymb}
\usepackage{graphics}
\usepackage{enumerate}
\usepackage{multicol}
\usepackage{epsfig}
\usepackage{csquotes}
\usepackage{enumitem}

\newtheorem{tr}{TR Item}
\renewcommand{\thetr}{\Alph{tr}} % "letter-numbered" theorems
\newtheorem{trexam}[tr]{TR Example}
\newtheorem{trthm}[tr]{TR Theorem}
\newtheorem{trlem}[tr]{TR Lemma}
\newtheorem{trdef}[tr]{TR Definition}

\newtheorem{thm}{Theorem}                                                       
\newtheorem{cor}[thm]{Corollary}                                                
\newtheorem{lem}[thm]{Lemma}                                                    
\newtheorem{prop}[thm]{Proposition}                                             
\newtheorem{ax}[thm]{Axiom}                                                     
\theoremstyle{definition}                                                       
\newtheorem{defn}[thm]{Definition}                                              
\newtheorem{exam}[thm]{Example}                                                 
\newtheorem{rem}[thm]{Remark} 
%\renewcommand*{\proofname}{Proof Sketch}

%
% For type setting inference rules with labels.
%
\newcommand{\inferlbl}[3]
			{\inferrule{#3}{#2}{\textsf{\footnotesize{\sc #1}}}}
\newcommand{\inferline}[3]
			{\inferrule{#3}{#2} & {\textsf{\footnotesize{\sc #1}}} \\ \\}

\newcommand{\Lagr}{\mathcal{L}}
\newcommand{\lang}[1]{\Lagr\{#1\}}
\newcommand{\stru}[2]{ {\tt string\_union}(#1,#2)}

\newcommand{\dconvert}[2]{ {\tt dconvert}(#1,#2) }
\newcommand{\filter}[2]{ {\tt filter}(#1,#2) }
\newcommand{\ifilter}[2]{ {\tt ifilter}(#1,#2) }

\newcommand{\reduces}{ \Rightarrow }
\newcommand{\gvd}{\Gamma \vdash }
\newcommand{\ovd}{\Omega \vdash }

\newcommand{\trep}{{\tt rep}}

\newcommand{\tstrf}[1]{`#1\textrm'} %??
\newcommand{\strf}[1]{``#1"}

\newcommand{\iso}{\cong}
%%
%% Source and Target language definitions.
%%
\newcommand{\lambdas}{\lambda_{RS}}
\newcommand{\lambdap}{\lambda_P}

% Source language terms.
\newcommand{\sisubst}[3]{{\sf rreplace}[#1](#2;#3)} \newcommand{\rreplace}[3]{{\sf rreplace}[#1](#2;#3)} % lots of legacy naming around these parts...

\newcommand{\rssreplace}[3]{\sisubst{#1}{#2}{#2}} % TODO-nrf fix this.

\newcommand{\coerce}[2]{ {\sf rcoerce}[#1](#2)}
\newcommand{\rcoerce}[2]{{\sf rcoerce}[#1](#2)}
\newcommand{\sistr}[1]{{\sf rstr}[#1]}   \newcommand{\rstr}[1]{{\sf rstr}[#1]} % Lots of legacy naming around these parts...

\newcommand{\val}{{\sf val}}

\newcommand{\rcheck}[4]{ {\sf rcheck}[#1](#2;#3;#4) }

\newcommand{\strin}[1]{\sistr{#1}}
\newcommand{\rsconcat}[2]{{\sf rconcat}(#1;#2)} \newcommand{\rconcat}[2]{{\sf rconcat}(#1;#2)} % lots of legact naming around these parts..

% Source language types.
\newcommand{\stringin}[1]{{\sf stringin}[#1]}

% target language terms.
\newcommand{\tsubst}[3]{{\sf replace}(#1;#2;#3)} \newcommand{\metareplace}[3]{{\sf replace}(#1;#2;#3)} % TODO-nrf rename the commands. Lots of legacy naming around these parts...

\newcommand{\tcheck}[4]{{\sf check}(#1; #2; #3; #4)}
\renewcommand{\tstr}[1]{{{\sf str}[#1]}}
\newcommand{\preplace}[3]{{\sf replace}(#1;#2;#3)}
\newcommand{\tconcat}[2]{{\sf concat}(#1;#2)} \newcommand{\concat}[2]{{\sf concat}(#1;#2)} % lots of legacy naming around these parts...

\newcommand{\regext}[1]{ {\sf rx}[#1] } % TODO-nrf remove
\newcommand{\rx}[1]{ {\sf rx}[#1] }

% Target language types
\newcommand{\str}{{\sf string}}
\newcommand{\regex}{{\sf regex}}

% Meta-theoretic functions
\newcommand{\lsubst}[3]{{\sf subst}(#1;#2;#3)} % This used to renderlreplace(...) so there're probably mistkes wherever this command was used now.
\newcommand{\lreplace}[3]{{\sf lreplace}(#1; #2; #3)}

\newcommand{\sctx}{\Psi} % Context for external typing
\newcommand{\tctx}{\Theta} % Context for internal typing
\newcommand{\ereduces}{\Downarrow}

\newcommand{\strcase}[3]{ {\sf rstrcase}(#1; #2; #3)}
\newcommand{\pstrcase}[3]{ {\sf strcase}(#1; #2; #3)}

\newcommand{\lhead}[1]{ {\sf lhead}(#1) }
\newcommand{\ltail}[1]{ {\sf ltail}(#1) }

% Judgements
\newcommand{\trden}[1]{\llbracket #1 \rrbracket} % = Translation Denotation.

% Relations
\newcommand{\treduces}{ \Downarrow }
\newcommand{\sreduces}{ \Downarrow }

%%
%% Constrain the size of full-page diagrams and rule lists
%%
%%\newcommand{\pagewidth}{5in}
%%\newcommand{\rulelistwidth}{3in}

% Names of type systems presented in paper
\newcommand{\lcs}{\lambda_{S}}

\setlength{\grammarindent}{3em}

% Arguments to lreplace are as follows:
% lreplace(pattern to replace, target, replacement).
% 
% ie lreplace(a,b,c) reads "replace all instances of a in b with c" and
% correspondes to Python's b.replace(a,c). 
\renewcommand{\lreplace}[3]{{\tt lreplace}(#1, #2, #3)}
\newcommand{\freplace}[3]{{\tt freplace}(#1, #2, #3)}
\newcommand{\freplacep}[4]{{\tt freplace}(#1,#2,#3,#4)}

\newcommand{\eps}{\epsilon}

\title{Statically Typed String Sanitation Inside a Python: Technical Report}
\author{Nathan Fulton \\ Cyrus Omar \\ Jonathan Aldrich}

\begin{document}
\linenumbers

\maketitle
\section{String and Language Replacement}

\subsection{The Trivial Definition}

\subsection{An Automaton Construction}

Insert Automaton stuff...

\subsection{Toward a Precise Definition}

\section{Proofs of Lemmas and Theorems about $\lambdas$}

This section presents proofs of lemmas and theorems about the type systems
presented in \cite{fulton2014}. In addition, we provide some examples to help 
motivate and explain definitions.

To facilitate the type safety proof, we introduce
a small step semantics for both $\lambdas$ and $\lambdap$. All theorems in
this section are proven as stated in \cite{fulton2014}.

Theorems and lemmas appearing in \cite{fulton2014} are numbered, while
supporting facts are lettered.

\subsection{Head and Tail Operations}

\begin{defn}[Definition of \lhead{r}]\label{def:lhead}
The relation $\lhead{r}=r'$ is defined in terms of the structure of $r$:
\begin{itemize}[noitemsep]
\item $\lhead{\epsilon} = \epsilon$
\item $\lhead{.} = a_1 + a_2 + ... + a_n$ for all $a_i \in \Sigma$ where $|\Sigma| = n$.
\item $\lhead{a} = a$ %where $a \in \Sigma$
\item $\lhead{r_1\cdot r_2} = \lhead{r_1}$
\item $\lhead{r_1 +  r_2}  = \lhead{r_1} + \lhead{r_2}$
\item $\lhead{r*} = \epsilon + \lhead{r}$
\end{itemize}
\end{defn}

\begin{defn}[Brzozowski's Derivative]\label{def:derivative}
  The \emph{derivative of $r$ with respect to $s$} is denoted by $\delta_s(r)$
  and is $\delta_s(r) = \{t | st \in \lang{r}\}$.
\end{defn}

\begin{defn}[Definition of \ltail{r}]
The relation $\ltail{r} = r'$ is defined in terms of $\lhead{r}$.
Note that $\lhead{r} = a_1 + a_2 + ... + a_i$.
We define $\ltail{r} = \delta_{a_1}(r) + \delta_{a_2}(r) + ... + \delta_{a_i}(r)$.
\end{defn}

\begin{trexam}[All the heads \\
of all the tails \\
can be more than \\
one head and tail]\label{exam:headsandtails}
$r \not = \lhead{r} \cdot \ltail{r}$.
\end{trexam}
\begin{proof}
A simple counter-example is $ab + cd$. 
Note that $\lhead{ab + cd} = a+c$ and $\ltail{ab + cd} = b + d$. 
Therefore, $\{ad,bc\} \subset \lang{\lhead{ab + cd} \cdot \ltail{ab + cd}}$ even though neither of these is in $\lang{r}$.
\end{proof}

\todo{clean up these paragraphs}
Example~\ref{exam:headsandtails} does not cause type soundness problems for ${\sf strcase}$ because $s \in \lang{r} \implies s \in \lhead{r}\cdot\ltail{r}$ is the
property required for soundness. Still, in a production implementation, it will make sense to massage the definitions of $\lhead{r}$ and $\ltail{r}$ so that type information is not unnecessarily lost during sub-string operations.

This is a general pattern in string operations -- $\lambdas$ simulates common operations on strings within the type system.
If there is an operation for concatenating to strings, we define an operation for concatenating to regular expressions.
If there is an operation for peeling off the first (n) characters of a string, then we define an operation for peeling off
the first (n) characters of a regular expression.

It is important to note, however, that we need not \emph{exactly} simulate the action of string operations using regular expressions.
In the case of concatenation, we lose some information -- of course the string ad is never in $r$, but after decomposing the string, our
type tells us this might actually be the case. That's okay, because the types are conservative in their approximation. Soundness is never violated.

In the case of string replacement, there are \emph{trivial} definitions of substitution (on strings) and replacement (on languages) which over-approximate 
the effect of a substitution. Closing these gaps in approximation is important, and motivates the string operations portion of this TR.

\subsection{Some Corollaries About Substitution and Language Replacement}

\begin{defn}[$\tt{subst}$]
We consider several choices in the previous section.
\end{defn}

\begin{defn}[$\tt{lreplace}$]
We consider several choices in the previous section.
\end{defn}

\begin{prop}[Closure] \label{thm:total}
  If $\lang{r}, \lang{r_1}$ and $\lang{r_2}$ are regular languages, then $\lang{\lreplace{r}{r_1}{r_2}}$ is also a regular language.
\end{prop}
\begin{proof}
Correctness implies closure.
\end{proof}

\begin{prop}[Substitution Correspondence] \label{thm:substcorrespondence}
  If $s_1 \in \lang{r_1}$ and $s_2 \in \lang{r_2}$ then $\lsubst{r}{s_1}{s_2} \in \lang{\lreplace{r}{r_1}{r_2}}$.
\end{prop}
\begin{proof}
This is exactly the correctness result proven for some pairs of ${\tt subst}$ and ${\tt replace}$ in the previous section.
\end{proof}

\begin{lem}[Properties of Regular Languages.] \label{thm:regexprops}
~
\begin{itemize}

\item If $s_1 \in \lang{r_1}$ and $s_2 \in \lang{r_2}$ then $s_1s_2 \in \lang{r_1\cdot r_2}$.
\item For all strings $s$ and regular expressions $r$, either $s \in \lang{r}$ or $s \not \in \lang{r}$.
\item Regular languages are closed under reversal.
\end{itemize}
\end{lem}

Lemma~\ref{thm:regexprops} states some well-known properties about regular expressions.

\begin{lem}
  If $\emptyset \vdash e : \stringin{r}$ then $r$ is a well-formed regular expression.
\end{lem}
\begin{proof}
  The proof proceeds by induction on the derivation of the premise.
  The only non-trivial cases (those which require more than an appeal to inversion) are S-T-Case, S-T-Replace and S-T-Concat.

In the S-T-Case case, note that ${\sf lhead}$ and ${\sf ltail}$ are total functions for well-formed regular expressions to well-formed regular expressions.

In the S-T-Concat case, note that \ref{thm:total} implies that if $r_1$ and $r_2$ are regular expressions then so is $r_1 \cdot r_2$.

In the S-T-Replace case, it suffices to show that $\lreplace{r}{r_1}{r_2}$ is a regular expression assuming (inductively) that $r, r_1$ and $r_2$ are all regular expressions.
This follows from the Closure proposition.
\end{proof}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Type Safety and Security.
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{The Small Step Semantics}

To prove type safety and the security theorem for the big step semantics, we
first prove type safety for a small step semantics in Figure~\ref{sdlambdas}
and then extend this to the big step semantics in Figure~\ref{dlambdas} by
proving a correspondence between the semantics.

%\newcommand{\therel}{R_{\stringin{r}}}
%\begin{trdef}[Relation over $\stringin{r}$]
%$\therel(e)$ iff $e$ halts and $\emptyset \vdash e : \stringin{r}$.
%\end{trdef}

Note that small step evaluation for $\lambdas$ is terminating:
%\begin{trthm}[Termination of small step semantics]
if $\emptyset \vdash e : \sigma$ then $e \mapsto^* v$ such that $v \ \val$.
%\end{trthm}
%\begin{proof}
We do not develop the full proof here, but note that the simply typed lambda calculus terminates.
For the string fragment, observe that the S-T- rules do not add any non-trivial binding structure because substitutions
$[e/x]e'$ may only occur in the special case where $e = \rstr{s}$, so that the length of the term
never increases and the number of free variables strictly decreases. Therefore, the standard normalization
argument proceeds without complication after fixing an evaluation order for the compatibility rules 
(all our other proofs are agnostic to evaluation order).
%\end{proof}

\begin{trlem}[Canonical Forms]\label{cfgen}
Suppose $v \ \val$.%or $e \sreduces v$.

If $\emptyset\vdash v : \stringin{r}$ then $v = \rstr{s}$.

If $\emptyset\vdash v : \sigma \rightarrow \sigma'$ then $v = \lambda x . e'$ for some $e'$.
\end{trlem}
\begin{proof}
By inspection of valuation and typing rules.
\end{proof}

For the sake of completeness, we include a statement of the weaker lemma
stated in the paper:

\begin{lem}[Canonical Forms for the String Fragment of $\lambdas$]\label{thm:cfs}
  If $\emptyset \vdash e : \stringin{r}$ and $e \sreduces v$ then $v=\rstr{s}$. 
\end{lem}
\begin{proof}
This fact follows directly from Lemma~\ref{cfgen}.
\end{proof}

\begin{trlem}[Progress of small step semantics.]
If $\vdash e : \sigma$ either $e \ \val$ or $e \mapsto e'$ for some $e'$.
\end{trlem}
\begin{proof}
The proof proceeds by induction on the derivation of $\vdash e : \sigma$.
\begin{itemize}[label=$ $,itemsep=1ex]
\item \textbf{$\lambda$ fragment}. Cases SS-T-Var, SS-T-Abs, and SS-T-App are exactly as in a proof of type safety for
the simply typed lambda calculus.

\item \textbf{S-T-Stringin-I}. Suppose $\vdash \rstr{s} : \stringin{s}$.
The $\rstr{s} \ \val$ by SS-E-RStr.

\item \textbf{S-T-Concat}. Suppose $\vdash \rconcat{e_1}{e_2} : \stringin{s}$.
By inversion and induction, $e_1 \mapsto e_1'$ or $e_1 \ \val$ and similarly
for $e_2$. 
If $e_1$ steps, then SS-E-Concat-Left applies and so
$\rconcat{e_1}{e_2} \mapsto \rconcat{e_1'}{e_2}$. Similarly, if $e_2$ steps
then $e$ steps by SS-E-Concat-Right.

In the remaining case, $e_1 \ \val$ and $e_2 \ \val$.
But then it follows by canonical forms that $e_1 = \rstr{s_1}$
and $e_2 = \rstr{s_2}$.
Note that $\rconcat{\rstr{s_1}}{\rstr{s_2}} \mapsto \rstr{s_1s_2}$ by SS-E-Concat.

\def\mline#1{\par\hspace*{-\leftmargin}\parbox{\textwidth}{\[#1\]}}
\item \textbf{S-T-Case}.
Suppose $e = \strcase{e_1}{e_2}{x,y.e_3}$.
By inversion, $e_1 : \stringin{r}$.
%\begin{align}
%&\vdash e_1 : \stringin{r} \\
%&\vdash e_2 : \sigma \\
%x:\stringin{\lhead{r}}, y : \stringin{\ltail{r}} &\vdash e_3 : \sigma
%\end{align}
%Actually this isn't even necessary...
%By (2), $e_2 \mapsto e_2'$ or $e_2 \val$. 
%In the former case, $e \mapsto \strcase{e_1}{e_2'}{x,y.e_3}$.
From this fact, induction, and canonical it follows that $e_1 \mapsto e_1'$ or $e_1 = \rstr{s}$.
In the former case, $e$ steps by S-E-Case-Left.
In the latter case, note that $s = \epsilon$ or $s = at$ for some string $t$.
If $s = \epsilon$ then $e$ steps by S-E-Case-$\epsilon$-Val,
and if $s = at$ the $e$ steps by S-E-Case-Concat.

\item \textbf{S-T-Replace}.
Suppose $e = \rreplace{r}{e_1}{e_2}$ and $e : \stringin{r'}$ where, by inversion of S-T-Replace,
\begin{align}
&\vdash e_1 : \stringin{r_1} \\
&\vdash e_2 : \stringin{r_2} \\
&\lreplace{r}{r_1}{r_2} = r'
\end{align}

By (1), inversion and induction $e_1 \ \val$ or $e_1 \mapsto e_1'$ for some $e_1'$,
If $e_1 \mapsto e_1'$ then $e$ steps by SS-E-Replace-Left.
Similarly, if $e_2$ steps then $e$ steps by SS-E-Replace-Right.
The only remaining case is where $e_1 \ \val$ and also $e_2 \ \val$.
But then by canonical forms, $e_1 = \strin{s_1}$ and $e_2 = \strin{s_2}$.
In this case, $e \mapsto \rstr{\subst{r}}{s_1}{s_2}$ by SS-E-Replace.

\item \textbf{S-T-SafeCoerce}.
Suppose that $\vdash \rcoerce{r}{e_1} : \stringin{r}$.
By inversion of S-T-SafeCoerce,
%\begin{align}
$\vdash e_1 : \stringin{r'}$% \\
%&\lang{r'} \subseteq \lang{r}
%\end{align}
By induction, $e_1 \ \val$ or $e_1 \mapsto e_1'$ for some $e_1'$.
If $e_1 \mapsto e_1'$ then $e$ steps by SS-E-SafeCoerce-Step.
Otherwise, $e_1 \ \val$ and by canonical forms $e_1 = \rstr{s}$.
In this case, $e = \rcoerce{r}{\rstr{s}} \mapsto \rstr{s}$ by SS-E-SafeCoerce.

\item \textbf{S-T-SafeCheck}
Suppose that $\vdash \rcheck{r}{e_1} : \stringin{r}$.
By inversion of S-T-Check:
\begin{align}
&\vdash e_0 : \stringin{r_0}  \\
x:\stringin{r} &\vdash e_1 : \sigma \\
&\vdash e_2 : \sigma \label{align6}
\end{align}

By (\ref{align6}) and induction,  $e_0 \mapsto e_0'$ or $e_0 \ \val$.
In the former case $e$ steps by SS-E-Check-StepRight.
Otherwise, by canonical forms $e_0 = \rstr{s}$.
By Lemma~\ref{thm:regexprops}, $s \in \lang{r_0}$ or else not.
In either case, the SS-E-Check-Ok and SS-E-Check-NotOk steps $e$.

\end{itemize}

\end{proof}

\begin{trlem}[Preservation for Small Step Semantics]
If $\emptyset \vdash e : \sigma$ and $e \mapsto e'$ then $\emptyset \vdash e : \sigma$.
\end{trlem}
\begin{proof}

By induction on the derivation of $e \mapsto e'$.

\begin{itemize}[label=$ $,itemsep=1ex]
\item \textbf{$\lambda$ fragment}. Cases SS-T-Var, SS-T-Abs, and SS-T-App are exactly as in a proof of type safety for
the simply typed lambda calculus.

\item \textbf{SS-E-Concat-Left}.
Suppose $e = \rconcat{e_1}{e_1} \mapsto \rconcat{e_1'}{e_2}$.
By inversion of S-T-Concat, $\vdash e_1 : \stringin{r_1}$ where $\vdash e : \stringin{r_1r_2}$.
By induction if $e_1 \mapsto e_1'$ then $\vdash e_1' : \stringin{r_1}$.
Therefore, $\vdash \rconcat{e_1'}{e_2} : \stringin{r_1r_2}$.

\item {\textbf{SS-E-Concat-Right}}. Similar to SS-E-Concat-Left.

\item \textbf{SS-E-Concat}. Suppose $\rconcat{\rstr{s_1}}{\rstr{s_2}} : \stringin{r_1r_2}$ and
$\rconcat{\rstr{s_1}}{\rstr{s_2}} \mapsto \rstr{s_1s_2}$.
Then by inversion $\rstr{s_1} : \stringin{r_1}$ and similarly for $\rstr{s_2}$.
Therefore, $s_1 \in \lang{r_1}$ and $s_2 \in  \lang{r_2}$ from which it follows by Lemma~\ref{thm:regexprops}
that $s_1s_2 \in \lang{r_1r_2}$. Therefore, $\vdash \rstr{s_1s_2} : \stringin{r_1r_2}$ by S-T-Rstr.

\item \textbf{S-E-Case-Left}.
Suppose that $e = \strcase{e_1}{e_2}{x,y.e_3} \mapsto \strcase{e_1'}{e_2}{x,y.e_3}$
and $\emptyset \vdash e : \stringin{r}$.
By inversion of S-T-Case:
\begin{align}
&\vdash e_1 : \stringin{r} \label{align7} \\
&\vdash e_2 : \sigma \label{align8} \\ 
x:\stringin{\lhead{r}}, y : \stringin{\ltail{r}} &\vdash e_3 : \sigma \label{align9}
\end{align}
By (\ref{align7}) and the assumption that $e_1 \mapsto e_1'$, it follows by induction that $e_1' : \stringin{r}$.
This fact together with (\ref{align8}) and (\ref{align9}) implies by S-T-Case that
$\strcase{e_1'}{e_2}{x,y.e_3} : \sigma$.

\item \textbf{SS-E-Case-Right}. 
Suppose that $e = \strcase{e_1}{e_2}{x,y.e_3} \mapsto \strcase{e_1}{e_2'}{x,y.e_3}$
and $\emptyset \vdash e : \stringin{r}$.
By inversion of S-T-Case:
\begin{align}
&\vdash e_1 : \stringin{r} \label{align10}\\
&\vdash e_2 : \sigma \label{align11} \\ 
x:\stringin{\lhead{r}}, y : \stringin{\ltail{r}} &\vdash e_3 : \sigma \label{align12}
\end{align}
By (\ref{align11}) and the assumption that $e_2 \mapsto e_2'$, it follows by induction that $e_2' : \stringin{r}$.
This fact together with (\ref{align10}) and (\ref{align12}) implies by S-T-Case that 
$\strcase{e_1}{e_2'}{x,y.e_3} : \sigma$.

\item \textbf{SS-E-Case-$\epsilon$-Val}.
Suppose $e = \strcase{-}{e_2}{-}: \sigma$ and $e \mapsto e_2$.
By inversion of S-T-Case, $e_2 : \sigma$.

\item \textbf{SS-E-Case-Concat}.
Suppose that $e = \strcase{\rstr{as}}{e_2}{x,y.e_3} \mapsto [\rstr{a}, \rstr{s} / x,y] e_3$
and that $e : \sigma$.
By inversion of S-T-Case:
\begin{align}
&\vdash \rstr{as} : \stringin{r} \label{align13}\\
&\vdash \rstr{e_2} : \sigma \label{align14} \\ 
x:\stringin{\lhead{r}}, y : \stringin{\ltail{r}} &\vdash e_3 : \sigma \label{align15}
\end{align}

We know that $as \in \lang{r}$ by (\ref{align13}) and inversion of S-T-Rstr.
Therefore, $a \in \lhead{r}$ by definition of ${\sf lhead}$.
Furthermore, $\ltail{r} = ... | \delta_a{r} | ...$ by definition of ${\sf ltail}$.
Note that $s \in \lang{\delta_a{r}}$ by definition of the derivative, and so
$s \in \ltail{r}$

From these facts about $a$ and $s$ we know by S-T-Rstr that $\vdash \rstr{a} : \stringin{\lhead{r}}$
and $\vdash \rstr{s} : \stringin{\lhead{r}}$.
It follows by (\ref{align15}) that  $\vdash [\rstr{a}, \rstr{s} / x,y] e_3 : \sigma$.

\item \textbf{Cases SS-E-Replace-Left, SS-E-Replace-Right, SS-E-Check-StepLeft, SS-E-SafeCoerce-Step, SS-E-Check-StepRight}.
\todo{At this point the method for handling compatibility cases is clear; therefore,
we elide these cases.}

\item \textbf{Case SS-E-Replace}.

Suppose $e = \rreplace{r}{\rstr{s_1}}{\rstr{s_2}} \mapsto \rstr{\lsubst{r}{s_1}{s_2}}$.
Assume $\emptyset \vdash e : \stringin{r'}$ for $r' = \lreplace{r}{r_1}{r_2}$.
Then by inversion of S-T-Replace:
\begin{align*}
\emptyset &\vdash \rstr{s_1} : \stringin{r_1} \\
\emptyset &\vdash \rstr{s_2} : \stringin{r_2}
\end{align*}
from which follows that $s_1 \in \lang{r_1}$ and $s_2 \in \lang{r_2}$.
Therefore, $\lsubst{r}{s_1}{s_2} \in \lang{\lreplace{r}{r_1}{r_2}}$ by Theorem~\ref{thm:substcorrespondence}.
It is finally derivable by S-T-Rstr that:

$\emptyset \vdash \rstr{\lsubst{r}{s_1}{s_2}} : \stringin{\lreplace{r}{r_1}{r_2}}$.

\item \textbf{Case SS-E-SafeCoerce}.
Suppose that $\rcoerce{r}{s_1} \mapsto \rstr{s_1}$ and that $\emptyset\vdash \rcoerce{r}{s_1} : \stringin{r}$.
By inversion of S-T-SafeCoerce we know that $s \in \lang{r}$ from which it follows by S-T-Rstr that $\emptyset\vdash s : \stringin{r}$.

\item \textbf{Case SS-E-Check-Ok}.
Suppose that $\rcheck{r}{\rstr{s}}{x.e_1}{e_2} \mapsto \subst{\rstr{s}}{x}{e_1}$,
$s \in \lang{r}$ and that $\emptyset \vdash \rcheck{r}{\rstr{s}}{x.e_1}{e_2} : \sigma$.
By inversion of S-T-Check, $x : \stringin{r} \vdash e_1 : \sigma$.
Note that $s \in \lang{r}$ implies that $s : \stringin{r}$ by S-T-RStr.
Therefore, $\emptyset \vdash \subst{\rstr{s}}{x}{e_1} : \sigma$

\item \textbf{Case SS-E-Check-NotOk}.
Suppose that $\rcheck{r}{\rstr{s}}{x.e_1}{e_2} \mapsto e_2$,
$s \not \in \lang{r}$ and that $\emptyset \vdash \rcheck{r}{\rstr{s}}{x.e_1}{e_2} : \sigma$.
By inversion of S-T-Check, $\emptyset \vdash e_2 : \sigma$.

\end{itemize}
\end{proof}

\begin{trthm}[Type Safety for small step semantics.] \label{thm:sstypesafte}
If $\emptyset\vdash e : \sigma$ then either $e \ \val$ or $e \mapsto^* e'$ and $\emptyset\vdash e' : \sigma$.
\end{trthm}
\begin{proof}
Follows directly from progress and preservation.
\end{proof}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% BEGIN THE SEMANTIC CORRESPONDENCE ARGUMENT FOR LAMBDA_RS
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsubsection{Semantic Correspondence between Big and Small Step Semantics for $\lambdas$}

Before extending these the type safety result fir the small step semantics to the big step semantics, we first establish a correspondence between
the big step semantics in Figure~\ref{fig:sdlambdas} and the small step semantics in Figure~\ref{fig:dlambdas}.
We prove the relevant theorems for the $\mapsto$ relation because these proofs are easier. We then extend this result via the correspondence theorem
to the more concise big step semantics presented in \cite{fulton2014}.

\begin{trthm}[Semantic Correspondence for $\lambdas$ (Part I)]
If $e \sreduces v$ then $e \mapsto^* v$.  
\end{trthm}
\begin{proof}
We proceed by structural induction on $e$.

\begin{itemize}[label=$ $,itemsep=1ex]

% Abstraction case.
\item \textbf{Case $e = \lambda x . e_1$}. 
The only applicable rule is S-E-Abs, so $v = \lambda x . e_1$.
Note that $\lambda x . e_2 \mapsto^* \lambda x . e_2$ by RT-Refl.

% Application case.
\item \textbf{Case $e=e_1(e_2)$}.
The only applicable rule is S-E-App. By inversion, we establish that the following:
\begin{align*}
	e_1 &\sreduces \lambda x.e_1' \\
	e_2 &\sreduces v_2 \\
	[v_2/x]e_1' &\sreduces v
\end{align*}

From which it follows by induction that:
\begin{align*}
	e_1 &\mapsto^* \lambda x.e_1'  \\
	e_2 &\mapsto^* v_2 \\
	[v_2/x]e_1' &\mapsto^* v
\end{align*}

Note that the following rule is derivable by repeating applications of the left and right compatibility rules for application:
\begin{mathpar}
\inferrule[L*-App]
{ e_1 \mapsto^* e_1' \\ e_2 \mapsto^* e_2'  }
{ e_1(e_2) \mapsto^* e_1'(e_2') }
\end{mathpar}

From these facts and L-AppAbs, we may establish that $e_1(e_2) \mapsto^* ( \lambda x.e_2 )(v_2) \mapsto [v_2/x]e_2$.
Note that $[v_2/x]e_2 \mapsto^* v$, so by RT-Trans it follows that $e = e_1(e_2) \mapsto^* v$.

% String Values
\item \textbf{Case $e = \strin{s}$}. The only applicable rule is S-E-RStr, so $v = \strin{s}$.
By RT-Refl, $\strin{s} \mapsto^* \strin{s}$.

% String Concatenation.
\item \textbf{Case $e=\rconcat{e_1}{e_2}$}. 
The only applicable rule is S-E-Concat, so $v = \rstr{s_1s_2}$.
By inversion, $e_1 \sreduces \strin{s_1}$ and $e_2 \sreduces \strin{s_2}$.
By induction, $e_1 \mapsto^* \strin{s_1}$ and $e_2 \mapsto^* \strin{s_2}$.
Note that the rule following is derivable:
\begin{mathpar}
\inferrule[SS-E-Concat-LR*]
{ e_1 \mapsto^* e_1' \\ e_2 \mapsto^* e_2'  }
{ \rconcat{e_1}{e_2} \mapsto^* \rconcat{e_1'}{e_2'} }
\end{mathpar}

From these facts, it follows that $\rconcat{e_1}{e_2} \mapsto^* \rconcat{\strin{s_1}}{\strin{s_2}}$.
Finally, $\rconcat{\strin{s_1}}{\strin{s_2}} \mapsto \strin{s_1s_2}$ by SS-E-Concat.
By RT-Step, it follows that $\rconcat{e_1}{e_2} \mapsto^* \strin{s_1s_2}$.

%Strcase Case.
\item \textbf{Case $e=\strcase{e_1}{e_2}{x,y.e_3}$}.

There are two subcases. For the first, suppose $\strcase{e_1}{e_2}{x,y.e_3} \sreduces v$ was finally derived by S-E-Case-$\epsilon$.
By inversion: 
\begin{align*}
e_1 &\sreduces \strin{\epsilon} \\
e_2 &\sreduces v
\end{align*}
from which it follows by induction that:
\begin{align*}
e_1 &\mapsto^* \strin{\epsilon} \\
e_2 &\mapsto^* v
\end{align*}

Note that the following rule is derivable:

\begin{mathpar}
\inferrule[SS-E-Case-LR*]
{ e_1 \mapsto^* e_1' \\ e_2 \mapsto^* e_2'  }
{ \strcase{e_1}{e_2}{x,y.e_3} \mapsto^* \strcase{e_1'}{e_2'}{x,y.e_3} }
\end{mathpar}

From these facts is follows that $e \mapsto^* \strcase{\strin{\epsilon}}{v}{x,y.e_3}$.
By S-E-Case-$\epsilon$-Val and RT-Step it follows that $e \mapsto^* v$.

Now consider the other case where $\strcase{e_1}{e_2}{x,y.e_3} \sreduces v$ was finally derived by S-E-Case-Concat.
By inversion, 
$e_1 \sreduces \strin{as}$ and $[\strin{a}, \strin{s} / x, y]e_3 \sreduces v$. 
From these facts it follows by induction that 
$e_1 \mapsto^* \strin{as}$ and
$[\strin{a}, \strin{s} / x, y]e_3 \mapsto^* v$.

By the first of these facts, it is derivable via SS-E-Case-LR* that 
$e \mapsto^* \strcase{e_1'}{\strin{as}}{x,y.e_3}$.
SE-E-Case-Concat applies to this form, so by RT-Step we know $e \mapsto^* [\strin{a}, \strin{s} / x, y]e_3$.
Recall that $[\strin{a}, \strin{s} / x, y]e_3 \mapsto^* v$, so by RT-Trans we finally derive $e \mapsto^* v$.

% String Replacement.
\item \textbf{Case $e=\rreplace{r}{e_1}{e_2}$}.
There is only one applicable rule, so $v = \rstr{s}$ and by inversion it follows that:
\begin{align*}
e_1 &\sreduces \rstr{s_1} \\
e_2 &\sreduces \rstr{s_2}
\end{align*}

From which it follows by induction that:
\begin{align*}
e_1 &\mapsto^* \rstr{s_1} \\
e_2 &\mapsto^* \rstr{s_2}
\end{align*}

Furthermore, $\lsubst{r}{s_1}{s_2} = s$ by induction.
Note that the following rule is derivable:

\begin{mathpar}
\inferrule[SS-E-Replace-LR*]
{ e_1 \mapsto^* e_1' \\ e_2 \mapsto^* e_2'  }
{ \rreplace{r}{e_1}{e_2} \mapsto^* \rreplace{r}{e_1'}{e_2'} }
\end{mathpar}

From these facts, $\rreplace{r}{e_1}{e_2} \mapsto^* \rreplace{r}{\rstr{s_1}}{\rstr{s_2}}$.

Finally, $\rreplace{r}{\rstr{s_1}}{\rstr{s_2}} \mapsto \lsubst{r}{s_1}{s_2}$.

From these two facts we know via TR-Step that $\rreplace{r}{e_1}{e_2} \mapsto^* \rreplace{r}{e_1}{e_2}$.
Recall that $\lsubst{r}{s_1}{s_2} = s$, from which the conclusion follows.

\item \textbf{Case $e = \rcoerce{r}{e_1}$}.
In this case $e \sreduces v$ is only finally derivable via S-E-SafeCoerce.
Therefore, $v = \strin{s}$ and by inversion $e_1 \sreduces \strin{s}$.
By induction, $e_1 \mapsto^* \strin{s}$.

The following rule is derivable:

\begin{mathpar}
\inferrule[SS-E-SafeCoerce-Step]
{ e \mapsto^* e' }
{ \rcoerce{r}{e} \mapsto^* \rcoerce{r}{e'} }
\end{mathpar}

Applying this rule at $e_1 \mapsto^* \strin{s}$ derives $\rcoerce{r}{e_1} \mapsto^* \rcoerce{r}{\strin{s}}$.
In the final step, $\rcoerce{r}{\strin{s}} \mapsto \strin{s}$ by SS-E-SafeCoerce.
From this fact, we may derive via RT-Trans that $e \mapsto^* \strin{s}$ as required.

\item \textbf{Case $e = \rcheck{r}{e_1}{x.e_2}{e_3}$}.

Note that the rule following is derivable:

\begin{mathpar}
\inferrule[SS-E-Check-Step]
{ e_1 \mapsto^* e_1' \\ e_3 \mapsto^* e_3' }
{ \rcheck{r}{e_1}{x.e_2}{e_3} \mapsto^* \rcheck{r}{e_1'}{x.e_2}{e_3'} }
\end{mathpar}

There are two ways to finally derive $e \sreduces v$.
In both cases, $e_1 \sreduces \strin{s}$ by inversion.
Therefore, in both cases, $e_1 \mapsto^* \strin{s}$ by induction
and so $e \mapsto^* \rcheck{r}{\strin{s}}{x.e_2}{e_3}$ by SS-E-Check-Step.

Suppose $e \sreduces v$ is finally derived via SS-E-Check-Ok.
By the facts mentioned above and SS-E-Check-Step, $e \mapsto^* \rcheck{r}{\strin{s}}{x.e_2}{e_2}$.
Note that by inversion $s \in \lang{r}$. Therefore, SS-E-Check-Ok applies and so by RT-Trans $e \mapsto^* [\rstr{s}/x]e_1$.
By inversion, $[\strin{s}/x]e_1 \sreduces v$. Therefore, by induction and RT-Step $e \mapsto^* v$ as required.

Suppose that $e \sreduces v$ is instead finally derived via SS-E-Check-NotOk.
By inversion, $e_3 \sreduces v$ and by induction $e_3 \mapsto^* v$.
From these facts at SS-E-Check-Step, it is derivable that $e \mapsto^* \rcheck{r}{\strin{s}}{x.e_2}{v}$.

Also by inversion, $s \not \in \lang{r}$ and so SS-E-Check-NotOk applies.
Therefore, $\rcheck{r}{\strin{s}}{x.e_2}{v} \mapsto v$.

The conclusion $e \mapsto^* v$ follows from these facts by RT-Step.
\end{itemize}
\end{proof}
%% Abstraction case.
%\item \textbf{Case $e = \lambda x . e_1$}. 
%% String Values
%\item \textbf{Case $e = \strin{s}$}. 
%% String Concatenation.
%\item \textbf{Case $e=\rconcat{e_1}{e_2}$}. 
%%Strcase Case.
%\item \textbf{Case $e=\strcase{e_1}{e_1}{x,y.e_3}$}.
%% String Replacement.
%\item \textbf{Case $e=\rreplace{r}{e_1}{e_2}$}.
%% Coercion
%\item \textbf{Case $e = \rcoerce{r}{e_1}$}.
%% Replacement
%\item \textbf{Case $e = \rcheck{r}{e_1}{x.e_2}{e_3}$}.

Establishing the other direction requires a minor lemma about the relationship between values and reflexivity in the big step semantics.

\begin{trthm}[Small Step Values are Reflexive in the Big Step Semantics]
If $v \ \val$ in the small step semantics then $v \sreduces v$.
\end{trthm}
\begin{proof}
If $v \ \val$ is derived from SS-E-RStr then S-E-RStr establishes the result.
Otherwise $v \ \val$ is derived from SS-E-Abs and so S-E-Abs establishes the result.
\end{proof}

\begin{trthm}[Semantic Correspondence for $\lambdas$ (Part II)]
If $e \vdash \tau$, $e \mapsto^* v$ and $v \ \val$ then $e \sreduces v$.  
\end{trthm}
\begin{proof}
The proof proceeds by structural induction on $e$.

\begin{itemize}[label=$ $,itemsep=1ex]

\item \textbf{Case $e = \concat{e_1}{e_2}$}.
By inversion, $\vdash e_1 : \stringin{r_1}$.
By Type Safety, canonical forms and termination it follows that $e_1 \mapsto^* \rstr{s_1}$ for some $s_1$.
By induction, $e_1 \sreduces \rstr{s_1}$.

Similarly, $e_2 \mapsto^* \rstr{s_2}$ and $e_2 \sreduces \rstr{s_2}$.

Note that $\concat{e_1}{e_2} \mapsto^* \concat{\rstr{s_1}}{\rstr{s_2}} \mapsto \rstr{s_1s_2}$ by SS-E-Concat-LR* and S-E-Concat,
and so $e \mapsto^* \rstr{s_1s_2}$ by TR-Step. So it suffices to show that $e \sreduces \rstr{s_1s_2}$.

Finally, $e \sreduces \rstr{s_1s_2}$ follows via S-E-Concat from the facts that $e_1 \sreduces \rstr{s_1}$ and $e_2 \sreduces \rstr{s_2}$.
This completes the case.

\item \textbf{Case $e = \rreplace{r}{e_1}{e_2}$}.
By inversion of S-T-Replace, $\vdash e_1 : \stringin{r_1}$ for some $r_1$.
It follows by type safety, termination and canonical forms that $e_1 \mapsto^* \rstr{s_1}$.
By induction, $e_1 \sreduces \rstr{s_1}$.

Similarly, $e_2 \mapsto^* \rstr{s_2}$ and $e_2 \sreduces \rstr{s_2}$.

Note that $e \mapsto^* \rreplace{r}{\rstr{s_1}}{\rstr{s_2}} \mapsto \rstr{\lsubst{r}{s_1}{s_2}}$ by SS-Replace-LR* and SS-E-Replace
and so $e \mapsto^* \rstr{\lsubst{r}{s_1}{s_2}}$ by TR-Step.

It suffices to show $e \sreduces \rstr{\lsubst{r}{s_1}{s_2}}$, which follows by S-E-Replace from the facts that $e_1 \sreduces \rstr{s_1}$ and
similarly for $e_2 \sreduces \rstr{s_2}$.

\item \textbf{Case $e = \strcase{e_1}{e_2}{x.y.e_3}$}.
By inversion, $\vdash e_1 : \stringin{r}$ and $e_2 : \sigma$.
By type safety, canonical forms and termination $e_1 \mapsto^* \stringin{s_1}$ and by induction $e_1 \sreduces \stringin{s_1}$.
Similarly, $e_2 \mapsto^* v_2$ and $\vdash e_2 \sreduces v_2$.

By SS-E-Case-LR*, $\strcase{e_1}{e_2}{x.y.e_1} \mapsto^* \strcase{v_1}{v_2}{x.y.e_2}$.

Note that either $s_1 = \epsilon$ or $s_1 = as$ because we define strings as either empty or finite sequences of characters.
We proceed by cases.

If $s_1 = \epsilon$ then $\strcase{v}{v_2}{x,y.e_3} \mapsto v_2$ by SS-E-Case-$\epsilon$.
Therefore, by RT-Step, $e \mapsto^* v_2$.
Also, $e_1 \mapsto \strin{\epsilon}$ and $e_2 \mapsto v_2$ is enough to establish by S-E-Case-$\epsilon$ that $e \sreduces v_2$.

If $s_1 = as$ instead, then $\strcase{\strin{s_1}}{v_2}{x,y.e_3} \mapsto [\strin{a},\strin{s} / x,y ] e_3$ by SS-E-Case-Concat.
Inversion of the typing relation satisfies the assumptions necessary to appeal to termination.
Therefore,

%\todo{Note that this substitution does not modify the syntactic structure of $e_3$ except by replacing free variables wit strings, and so} by induction,
\[ [\strin{a},\strin{s} / x,y ] e_3 \mapsto^* v \text{ for } v \ \val. \]

It follows by RT-Step that $e \mapsto^* v$.

Note that the substitution does not change the structure of $e_3$.
So by induction, $[\strin{a},\strin{s} / x,y ] e_3 \sreduces v$.
Recall that $e_1 \sreduces \strin{s_1}$ and so by S-E-Case it follows that $e \sreduces [a,s / x,y] e_3 \sreduces v$.
\end{itemize}

\todo{The cases for coercion and checking are straightforward.}
\end{proof}

\subsection{Extension of Safety for Small Step Semantics}

\begin{thm}[Type Safety] \label{thm:typesafety}
  If $\emptyset \vdash e : \sigma$ 
  then $e \sreduces v$ and $\emptyset \vdash v : \sigma$.
\end{thm}
\begin{proof}
If $\emptyset \vdash e : \sigma$ then $e \mapsto^* v$ for $v \ \val$ by termination and type safety for the small step semantics.
Therefore, $e \sreduces v$ by part 2 of the semantic correspondence theorem.

Since $\emptyset \vdash e : \sigma$ and $e \mapsto^* v$, it follows that $\emptyset \vdash v : \sigma$ by type safety for the small step semantics.
\end{proof}

\subsubsection{The Security Theorem}\label{sec:securitythm}

\begin{thm}[Correctness of Input Sanitation for $\lambdas$]\label{thm:scorrect}
  If  $\emptyset \vdash e : \stringin{r}$ and $e \sreduces \rstr{s}$ then $s \in \lang{r}$.
\end{thm}
\begin{proof}
If $\emptyset \vdash e : \stringin{r}$ then $\emptyset \vdash \rstr{s} : \stringin{r}$ by type safety.
By inversion of S-T-Rstr, $s \in \lang{r}$.
\end{proof}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\renewcommand{\grammarlabel}[2]{#1\hfill#2}
\begin{figure}[b]
\small
  \begin{grammar}
<$r$> ::= $\epsilon$ | $.$ | $a$ | $r \cdot r$ | $r + r$ | $r*$ \hfill $a \in \Sigma$

\caption{Regular expressions over the alphabet $\Sigma$.}
\label{fig:regex}
\end{grammar}
\end{figure}
\begin{figure}[t]
\small
  \begin{grammar}
<$\sigma$> ::=  $\sigma \rightarrow \sigma$ | $\stringin{r}$    \hfill  source types

<$e$> ::= 
      $x$ | $\lambda x . e$ | $e(e) $\hfill source terms \alt 
      $\strin{s}$ | $\rsconcat{e}{e}$ | $\strcase{e}{e}{x,y.e}$ \hfill $s \in \Sigma^{*}$ \alt
      $\sisubst{r}{e}{e}$ | $\rcoerce{r}{e}$ | $\rcheck{r}{e}{x.e}{e}$

<$v$> ::= $\lambda x . e$ | $\strin{s}$ \hfill source values 
\caption{Syntax of $\lambda_{RS}$.}
\label{fig:glambdas}
\end{grammar}
\end{figure}

\renewcommand{\grammarlabel}[2]{#1\hfill#2}

\begin{figure}[t]
\small
  \begin{grammar}

<$\tau$> ::= $\tau \rightarrow \tau$ | $\str$ | $\regex$ \hfill target types

<$\iota$> ::= $x$ | $\lambda x . \iota$ | $\iota(\iota)$ \hfill target terms \alt
$\tstr{s}$ | $\tconcat{\iota}{\iota}$ | $\pstrcase{\iota}{\iota}{x,y.\iota}$ \alt
  $\rx{r}$ | $\preplace{\iota}{\iota}{\iota}$ | $\tcheck{\iota}{\iota}{\iota}{\iota}$

  <$\dot{v}$> ::= $\lambda x . \iota$ | \tstr{s} | \rx{r} \hfill target values

\end{grammar}
\caption{Syntax for the target language, $\lambdap$, containing strings and statically constructed regular expressions.}
\label{fig:lcsSyntax}
\end{figure}

\begin{figure}[t]
\small
$\fbox{\inferrule{}{\sctx \vdash e : \sigma}}$
~~~~$\sctx ::= \emptyset \pipe \sctx, x : \sigma$
\begin{mathpar}
\inferrule[S-T-Var]
{ x:\sigma \in \sctx }
{ \sctx \vdash x:\sigma}

\inferrule[S-T-Abs]
{\sctx, x : \sigma_1 \vdash e : \sigma_2}
{\sctx \vdash \lambda x.e : \sigma_1 \rightarrow \sigma_2}

\inferrule[S-T-App]
{\sctx \vdash e_1 : \sigma_2 \rightarrow \sigma \\ \sctx \vdash e_2 : \sigma_2}
{\sctx \vdash e_1(e_2) : \sigma}

\inferrule[S-T-Stringin-I]
{s \in \lang{r}}
{\sctx \vdash \strin{s} : \stringin{r}}

\inferrule[S-T-Concat]
{\sctx \vdash e_1 : \stringin{r_1} \\ \sctx \vdash e_2 : \stringin{r_2}}
{\sctx \vdash \rsconcat{e_1}{e_2} : \stringin{r_1 \cdot r_2}}

\inferrule[S-T-Case]
{ \sctx \vdash e_1 : \stringin{r} \\
  \sctx \vdash e_2 : \sigma \\
  \sctx, x : \stringin{\lhead{r}}, y : \stringin{\ltail{r}} \vdash e_3 : \sigma
}
{
  \sctx \vdash \strcase{e_1}{e_2}{x,y.e_3} : \sigma
}

\inferrule[S-T-Replace]
{\sctx \vdash e_1 : \stringin{r_1} \\ \sctx \vdash e_2 : \stringin{r_2} \\\\ \lreplace{r}{r_1}{r_2} = r'}
{\sctx \vdash \sisubst{r}{e_1}{e_2} : \stringin{r'}}

\inferrule[S-T-SafeCoerce]
{\sctx \vdash e : \stringin{r'} \\ \lang{r'} \subseteq \lang{r}}
{\sctx \vdash \coerce{r}{e} : \stringin{r}}

\inferrule[S-T-Check]
{\sctx \vdash e_0 : \stringin{r_0} \\ \sctx, x:\stringin{r} \vdash e_1 : \sigma \\ \sctx \vdash e_2 : \sigma}
{\sctx \vdash \rcheck{r}{e_0}{x.e_1}{e_2} : \sigma}
\end{mathpar}

\caption{Typing rules for $\lambdas$. 
The typing context $\sctx$ is standard.}
\label{fig:slambdas}
\end{figure}

\begin{figure}
  \small
$\fbox{\inferrule{}{e \sreduces v}}$
\begin{mathpar}
\inferrule[S-E-Abs]                                                             
{ \ }                                                                           
{\lambda x.e \sreduces \lambda x.e}

\inferrule[S-E-App]
{ e_1 \sreduces \lambda x . e_3 \\ e_2 \sreduces v_2 \\ [ v_2 / x ] e_3 \sreduces v }
{ e_1(e_2) \sreduces v }

\inferrule[S-E-RStr]
{ \ }
{\strin{s} \sreduces \strin{s}}

\inferrule[S-E-Concat]
{e_1 \sreduces \strin{s_1} \\ e_2 \sreduces \strin{s_2}} 
{\rsconcat{e_1}{e_2} \sreduces \strin{s_1 s_2}} % ???

\inferrule[S-E-Case-$\epsilon$]
{
  e_1 \sreduces \strin{\epsilon} \\
  e_2 \sreduces v_2
}
{
  \strcase{e_1}{e_2}{x,y.e_3} \sreduces v_2
}

\inferrule[S-E-Case-Concat]
{
  e_1 \sreduces \strin{a s} \\
  [\rstr{a},\rstr{s} / x,y]e_3 \sreduces v_3
}
{
  \strcase{e_1}{e_2}{x,y.e_3} \sreduces v_3
}

\inferrule[S-E-Replace]
{e_1 \sreduces \strin{s_1} \\ e_2 \sreduces \strin{s_2} \\ \lsubst{r}{s_1}{s_2} = s} 
{\sisubst{r}{e_1}{e_2} \sreduces \sistr{s}}

\inferrule[S-E-SafeCoerce]
{e \sreduces \strin{s}}
{\coerce{r}{e} \sreduces \strin{s}}

\inferrule[S-E-Check-Ok]
{e \sreduces \strin{s} \\ s \in \lang{r} \\ [\strin{s} / x]e_1 \sreduces v}
{
  \rcheck{r}{e}{x.e_1}{e_2} \sreduces v
}

\inferrule[S-E-Check-NotOk]
{
  e \sreduces \strin{s} \\ s \not \in \lang{r} \\
  e_2 \sreduces v
}
{
  \rcheck{r}{e}{x.e_1}{e_2} \sreduces v
}
\end{mathpar}
\caption{Big step semantics for $\lambdas$.}
\label{fig:dlambdas}
\end{figure}

\begin{figure}
  \small
$\fbox{\inferrule{}{e \ \val}}$

\begin{mathpar}
\inferrule[L-Val]
{  }
{\lambda x:\tau.t \ \val}
\end{mathpar}

$\fbox{\inferrule{}{e \mapsto e}}$

\begin{mathpar}

\inferrule[L-E-AppLeft]
{e_1 \mapsto e_1'}
{e_1 e_2 \mapsto e_1' e_2}

\inferrule[L-E-AppRight]
{e_2 \mapsto e_2'}
{e_1 e_2 \mapsto e_1 e_2'}

\inferrule[L-E-AppAbs]
{ \ }
{(\lambda x : \tau_{11} . t_{12}) v_2 \mapsto [v_2 / x]t_{12}}

\end{mathpar}

\fbox{\inferrule{}{e \mapsto^* e}}

\begin{mathpar}

\inferrule[RT-Refl]
{ \ }
{e \mapsto^* e}

\inferrule[RT-Trans]
{ e \mapsto^* e' \\ e' \mapsto^* e'' }
{ e \mapsto^* e'' }

\inferrule[RT-Step\footnote{This is a special case of RT-Trans which we distinguish in order to write clearer proofs.}]
{ e \mapsto^* e' \\ e' \mapsto v }
{ e \mapsto^* v }

\end{mathpar}

\label{fig:lambda}
\caption{Call-by-name small step Semantics for $\lambda$ and its reflexive, transitive closure.} \label{fig:lambda}
\end{figure}

\begin{figure}
  \small
%$\fbox{\inferrule{}{e \mapsto e}} \  \fbox{ \inferrule{}{v \ \val}}$
\begin{mathpar}
\inferrule[SS-E-RStr]
{ \ }
{\strin{s} \ \val}

\inferrule[SS-E-Concat-Left]
{e_1 \mapsto e_1'}
{\rsconcat{e_1}{e_2} \mapsto \rsconcat{e_1'}{e_2}}

\inferrule[SS-E-Concat-Right]
{e_2 \mapsto e_2'}
{\rsconcat{e_1}{e_2} \mapsto \rsconcat{e_1}{e_2'}}

\inferrule[SS-E-Concat]
{ \ }
{\rsconcat{\strin{s_1}}{\strin{s_2}} \mapsto \strin{s_1 s_2}} % ???

\inferrule[SS-E-Case-Left]
{
  e_1 \mapsto e_1'
}
{
  \strcase{e_1}{e_2}{x,y.e_3} \mapsto \strcase{e_1'}{e_2}{x,y.e_3}
}

\inferrule[SS-E-Case-Right]
{
  e_2 \mapsto e_2'
}
{
  \strcase{e_1}{e_2}{x,y.e_3} \mapsto \strcase{e_1}{e_2'}{x,y.e_3}
}

\inferrule[SS-E-Case-$\epsilon$-val]
{
  \ 
}
{
  \strcase{\strin{\epsilon}}{e_2}{x.y.e_3} \mapsto e_2
}

\inferrule[SS-E-Case-Concat]
{
  \ 
}
{
  \strcase{\strin{a s}}{e_2}{x,y.e_3} \mapsto [\rstr{a},\rstr{s} / x,y]e_3 
}

\inferrule[SS-E-Replace-Left]
{e_1 \mapsto e_1'}
{ \sisubst{r}{e_1}{e_2} \mapsto \sisubst{r}{e_1'}{e_2}}

\inferrule[SS-E-Replace-Right]
{e_2 \mapsto e_2'}
{ \sisubst{r}{e_1}{e_2} \mapsto \sisubst{r}{e_1}{e_2'}}

\inferrule[SS-E-Replace]
{ \ }
{\sisubst{r}{\strin{s_1}}{\strin{s_2}} \mapsto \sistr{\lsubst{r}{s_1}{s_2}}}

\inferrule[SS-E-SafeCoerce-Step]
{e \mapsto e'}
{\coerce{r}{e} \mapsto \coerce{r}{e'}}

\inferrule[SS-E-SafeCoerce]
{ \ }
{ \coerce{r}{\rstr{s}} \mapsto \rstr{s}}

%%% NRF-bookmark1
\inferrule[SS-E-Check-StepLeft]
{e \mapsto e'}
{\rcheck{r}{e}{x.e_1}{e_2} \mapsto \rcheck{r}{e'}{x.e_1}{e_2}}

\inferrule[SS-E-Check-StepRight]
{e_2 \mapsto e_2'}
{\rcheck{r}{e}{x.e_1}{e_2} \mapsto \rcheck{r}{e}{x.e_1}{e_2'}}

\inferrule[SS-E-Check-Ok]
{s \in \lang{r}}
{\rcheck{r}{\strin{s}}{x.e_1}{e_2} \mapsto [\rstr{s} / x]e_1}
\inferrule[SS-E-Check-NotOk]
{s \not \in \lang{r}}
{\rcheck{r}{\strin{s}}{x.e_1}{e_2} \mapsto e_2}

\end{mathpar}
\caption{Small step semantics for $\lambdas$. Extends \ref{fig:lambda}.}
\label{fig:dslambdas}
\end{figure}

\begin{figure}[t]\label{fig:lambdap}
\small
$\fbox{\inferrule{}{\tctx \vdash \iota : \tau}}$
~~~~$\tctx ::= \emptyset \pipe \tctx, x : \tau$

\begin{mathpar}
\inferrule[P-T-Var]
{ x:\tau \in \tctx }
{ \tctx \vdash x:\tau }

\inferrule[P-T-Abs]
{\tctx, x : \tau_1 \vdash \iota_2 : \tau_2}
{\tctx \vdash \lambda x.\iota_2 : \tau_1 \rightarrow \tau_2}

\inferrule[P-T-App]
{\tctx \vdash \iota_1 : \tau_2 \rightarrow \tau \\ \tctx \vdash \iota_2 : \tau_2}
{\tctx \vdash \iota_1(\iota_2) : \iota}

\inferrule[P-T-String]
{ \ }
{\tctx \vdash \tstr{s} : \str}

\inferrule[P-T-Regex]
{ \ }
{\tctx \vdash \rx{r} : \regex}

\inferrule[P-T-Concat]
{\tctx \vdash \iota_1 : \str \\ \tctx \vdash \iota_2 : \str}
{\tctx \vdash \tconcat{\iota_1}{\iota_2} : \str}

\inferrule[P-T-Case]
{\tctx \vdash \iota_1 : \str \\ \tctx \vdash \iota_2 : \tau \\ \tctx, x:\str, y:\str \vdash \iota_3:\tau}
{\tctx \vdash \pstrcase{\iota_1}{\iota_2}{x,y.\iota_3} : \tau}

\inferrule[P-T-Replace]
{\tctx \vdash \iota_1 : \regex \\ \tctx \vdash \iota_2 : \str \\ \tctx \vdash \iota_3 : \str }
{\tctx \vdash \preplace{\iota_1}{\iota_2}{\iota_3} : \str}

\inferrule[P-T-Check]
{\tctx \vdash \iota_r : \regex \\ \tctx \vdash \iota_1 : \str \\ \tctx \vdash \iota_2 : \sigma \\ \tctx \vdash \iota_3 : \sigma}
{\tctx \vdash \tcheck{\iota_r}{\iota_1}{\iota_2}{\iota_3} : \sigma}
\end{mathpar}
\caption{Typing rules for $\lambdap$.
The typing context $\tctx$ is standard.}
\label{fig:slambdap}
\end{figure}

\begin{figure}[t]
\small
$\fbox{\inferrule{}{\iota \treduces \dot{v}}}$

\begin{mathpar}
\inferrule[P-E-Abs]
{ \ }
{\lambda x.e \sreduces \lambda x.e}

\inferrule[P-E-App]
{ \iota_1 \sreduces \lambda x . \iota_3 \\  \iota_2 \sreduces \dot{v}_2 \\ [\dot{v}_2 / x] \iota_3 \sreduces \dot{v}_3}
{ \iota_1(\iota_2) \sreduces \dot{v}_3}

\inferrule[P-E-Str]
{ \ }
{\tstr{s} \treduces \tstr{s}}

\inferrule[P-E-Rx]
{ \ }
{\rx{r} \treduces \rx{r}}

\inferrule[P-E-Concat]
{\iota_1 \treduces \tstr{s_1} \\ \iota_2 \treduces \tstr{s_2}} 
{\tconcat{\iota_1}{\iota_2} \treduces \tstr{s_1 s_2}} % ???

\inferrule[P-E-Case-$\epsilon$]
{
  \iota_1 \treduces \tstr{ \epsilon } \\
  \iota_2 \treduces \dot{v_2} \\
}
{
  \pstrcase{\iota_1}{\iota_2}{x,y.\iota_3} \treduces \dot{v_2}
}

\inferrule[P-E-Case-Concat]
{
  \iota_1 \treduces \tstr{as} \\
  [\tstr{a},\tstr{s}/x,y] \iota_3 \treduces \dot{v} 
}
{
  \pstrcase{\iota_1}{\iota_2}{x,y.\iota_3} \treduces \dot{v}
}

\inferrule[P-E-Replace]
{\iota_1 \treduces \rx{r} \\ \iota_2 \treduces \tstr{s_2} \\ \iota_3 \treduces \tstr{s_3} \\ \lsubst{r}{s_2}{s_3} = s} 
{\preplace{\iota_1}{\iota_2}{\iota_3} \treduces \tstr{s}}

\inferrule[P-E-Check-OK]
{
  \iota_r \treduces \rx{r} \\
  \iota \treduces \tstr{s} \\ 
s \in \lang{r} \\ 
\iota_1 \treduces \dot{v_1} 
}
{
  \tcheck{\iota_r}{\iota}{\iota_1}{\iota_2} \treduces \dot{v_1}
}

\inferrule[P-E-Check-NotOK]
{\iota_r \treduces \rx{r} \\ \iota \treduces \tstr{s} \\ s \not \in \lang{r} \\ \iota_2 \treduces \dot{v_2}}
{\tcheck{\iota_r}{\iota}{\iota_1}{\iota_2} \treduces \dot{v_2}}
\end{mathpar}
\caption{Big step semantics for $\lambdap$}.
\label{fig:dlambdap}
\end{figure}

\begin{figure}[t]
\small
$\fbox{\inferrule{}{\iota \treduces \dot{v}}}$

\begin{mathpar}
\inferrule[SP-E-Abs]
{ \ }
{\lambda x.e \sreduces \lambda x.e}

\inferrule[SP-E-App]
{ \iota_1 \sreduces \lambda x . \iota_3 \\  \iota_2 \sreduces \dot{v}_2 \\ [\dot{v}_2 / x] \iota_3 \sreduces \dot{v}_3}
{ \iota_1(\iota_2) \sreduces \dot{v}_3}

\inferrule[SP-E-Str]
{ \ }
{\tstr{s} \treduces \tstr{s}}

\inferrule[SP-E-Rx]
{ \ }
{\rx{r} \treduces \rx{r}}

\inferrule[SP-E-Concat]
{\iota_1 \treduces \tstr{s_1} \\ \iota_2 \treduces \tstr{s_2}} 
{\tconcat{\iota_1}{\iota_2} \treduces \tstr{s_1 s_2}} % ???

\inferrule[SP-E-Case-$\epsilon$]
{
  \iota_1 \treduces \tstr{ \epsilon } \\
  \iota_2 \treduces \dot{v_2} \\
}
{
  \pstrcase{\iota_1}{\iota_2}{x,y.\iota_3} \treduces \dot{v_2}
}

\inferrule[SP-E-Case-Concat]
{
  \iota_1 \treduces \tstr{as} \\
  [\tstr{a},\tstr{s}/x,y] \iota_3 \treduces \dot{v} 
}
{
  \pstrcase{\iota_1}{\iota_2}{x,y.\iota_3} \treduces \dot{v}
}

\inferrule[SP-E-Replace]
{\iota_1 \treduces \rx{r} \\ \iota_2 \treduces \tstr{s_2} \\ \iota_3 \treduces \tstr{s_3} \\ \lsubst{r}{s_2}{s_3} = s} 
{\preplace{\iota_1}{\iota_2}{\iota_3} \treduces \tstr{s}}

\inferrule[SP-E-Check-OK]
{
  \iota_r \treduces \rx{r} \\
  \iota \treduces \tstr{s} \\ 
s \in \lang{r} \\ 
\iota_1 \treduces \dot{v_1} 
}
{
  \tcheck{\iota_r}{\iota}{\iota_1}{\iota_2} \treduces \dot{v_1}
}

\inferrule[SP-E-Check-NotOK]
{\iota_r \treduces \rx{r} \\ \iota \treduces \tstr{s} \\ s \not \in \lang{r} \\ \iota_2 \treduces \dot{v_2}}
{\tcheck{\iota_r}{\iota}{\iota_1}{\iota_2} \treduces \dot{v_2}}
\end{mathpar}
\caption{Small step semantics for $\lambdap$}.
\label{fig:sdlambdap}
\end{figure}

\begin{figure}[h]
\small
$\fbox{\inferrule{}{\trden{\sigma}=\tau}}$
\begin{mathpar}
\inferrule[Tr-T-String]
{ }
{ \trden{\stringin{r}} = \str }

\inferrule[Tr-T-Arrow]
{ \trden{\sigma_1} = \tau_1 \\ \trden{\sigma_2} = \tau_2 }
{ \trden{\sigma_1 \rightarrow \sigma_2} = \tau_1 \rightarrow \tau_2 }
\end{mathpar}
$\fbox{\inferrule{}{\trden{\Psi}=\Theta}}$
\begin{mathpar}
\inferrule[Tr-T-Context-Emp]
{ }{\trden{\emptyset} = \emptyset}

\inferrule[Tr-T-Context-Ext]
{\trden{\Psi}=\Theta\\ \trden{\sigma} = \tau}
{\trden{\Psi, x : \sigma} = \Theta, x : \tau}
\end{mathpar}
  $\fbox{\inferrule{}{ \trden{e} = \iota }}$
\begin{mathpar}
\inferrule[Tr-Var]
{ }
{ \trden{x} = x}

\inferrule[Tr-Abs]
{  \trden{e} = \iota }
{ \trden{ \lambda x.e } = \lambda x . \iota}

\inferrule[Tr-App]
{ \trden{e_1} = \iota_1 \\  \trden{e_2} = \iota_2}
{ \trden{e_1(e_2)} = \iota_1(\iota_2)}

\inferrule[Tr-Case]
{ \trden{e_1} = \iota_1 \\
   \trden{e_2} = \iota_2 \\
   \trden{e_3} = \iota_3
}
{ \trden{ \strcase{e_1}{e_2}{x,y.e_3} } = \pstrcase{\iota_1}{\iota_2}{x,y.\iota_3} }

\inferrule[Tr-string]
{ \ }
{   \trden{\strin{s}} = \tstr{s}}

\inferrule[Tr-Concat]
{  \trden{e_1} = \iota_1 \\  \trden{e_2} = \iota_2}
{  \trden{\rsconcat{e_1}{e_2}} = \tconcat{\iota_1}{\iota_2} }

\inferrule[Tr-Subst]
{   \trden{e_1} = \iota_1 \\   \trden{e_2} = \iota_2 }
{   \trden{ \sisubst{r}{e_1}{e_2} } = \tsubst{\rx{r}}{\iota_1}{\iota_2} }

\inferrule[Tr-SafeCoerce]
{  \trden{e} = \iota }
{   \trden{ \coerce{r'}{e} } = \iota }

\inferrule[Tr-Check]
{  
  \trden{e} = \iota \\  
   \trden{e_1} = \iota_1 \\  
   \trden{e_2} = \iota_2
} 
{ 
  \trden{ \rcheck{r}{e}{x.e_1}{e_2} } = \tcheck{\rx{r}}{\iota}{(\lambda x . \iota_1)(\iota)}{\iota_2} 
} 
\end{mathpar}
\caption{Translation from source terms ($e$) to target terms ($\iota$).}
\label{fig:tr}
\end{figure}

\bibliographystyle{abbrv}

\bibliography{research}

\end{document}

### 2014-10-16 sanitation-psp14/tr/strings_chapter.tex
Proof reading.

@@
 \usepackage{verbatimbox}
-\input{../att-icfp14/macros-atlam}
+\input{../../att-icfp14/macros-atlam}
 \usepackage{alltt}
@@
 \newtheorem{trdef}[tr]{TR Definition}
+\newtheorem{trconj}[tr]{TR Conjecture}
 
@@
 \maketitle
-\section{String and Language Replacement}
-
-\subsection{The Trivial Definition}
-
-\subsection{An Automaton Construction}
-
-Insert Automaton stuff...
-
-\subsection{Toward a Precise Definition}
+
+\begin{abstract}
+This Technical Report contains supporting materials for Statically Typed
+String Sanitation Inside a Python \cite{fulton2014} including proofs of lemmas
+and theorems asserted in the paper, examples, and additional discussion of
+the paper's technical content.
+\end{abstract}
 
@@
 This section presents proofs of lemmas and theorems about the type systems
-presented in \cite{fulton2014}. In addition, we provide some examples to help 
+presented in \cite{fulton2014}, the accompanying paper. In addition, we provide some examples to help 
 motivate and explain definitions.
@@
 Theorems and lemmas appearing in \cite{fulton2014} are numbered, while
-supporting facts are lettered.
+supporting facts appearing only in the Technical Report are lettered.
 
@@
 
-\begin{trexam}[All the heads \\
-of all the tails \\
-can be more than \\
-one head and tail]\label{exam:headsandtails}
+\begin{trexam}[All the heads of all the tails can be more than one head and tail]\label{exam:headsandtails}
 $r \not = \lhead{r} \cdot \ltail{r}$.
@@
 
-\todo{clean up these paragraphs}
-Example~\ref{exam:headsandtails} does not cause type soundness problems for ${\sf strcase}$ because $s \in \lang{r} \implies s \in \lhead{r}\cdot\ltail{r}$ is the
+Example~\ref{exam:headsandtails} does not imply a counter-example to type soundness because $s \in \lang{r} \implies s \in \lhead{r}\cdot\ltail{r}$ is the
 property required for soundness.
-Still, in a production implementation, it will make sense to massage the definitions of $\lhead{r}$ and $\ltail{r}$ so that type information is not unnecessarily lost during sub-string operations.
-
-This is a general pattern in string operations -- $\lambdas$ simulates common operations on strings within the type system.
-If there is an operation for concatenating to strings, we define an operation for concatenating to regular expressions.
-If there is an operation for peeling off the first (n) characters of a string, then we define an operation for peeling off
-the first (n) characters of a regular expression.
-
-It is important to note, however, that we need not \emph{exactly} simulate the action of string operations using regular expressions.
-In the case of concatenation, we lose some information -- of course the string ad is never in $r$, but after decomposing the string, our
-type tells us this might actually be the case. That's okay, because the types are conservative in their approximation. Soundness is never violated.
+Still, in a production implementation, it will make sense to massage the definitions of $\lhead{r}$ and $\ltail{r}$ so that type information is not unnecessarily lost during substring operations.
+
+This is a general pattern in string operations: $\lambdas$ simulates -- within the type system -- common operations on strings. 
+If there is an operation for concatenating to strings, we define an operation for concatenating two regular expressions.
+If there is an operation for peeling off the first (n) characters of a string, then we define an operation for 
+converting a regular expression $r$ into a regular expression $r'$ which recognizes any $n^{th}$ suffix of a string in $r$.
+
+It is important to note, however, that the type system need not \emph{exactly} simulate the action of string operations.
+In the case of concatenation, we lose some information because more string values are possible -- according the types --
+than are actually possible in the dynamic semantics.
+Soundness is not lost because the types are conservative in their approximation.
 
 In the case of string replacement, there are \emph{trivial} definitions of substitution (on strings) and replacement (on languages) which over-approximate 
-the effect of a substitution. Closing these gaps in approximation is important, and motivates the string operations portion of this TR.
+the effect of a substitution. Closing these gaps in approximation is important, and motivates the string operations portion of this technical report.
 
@@
 \begin{defn}[$\tt{subst}$]
-We consider several choices in the previous section.
+We consider several choices in the string operations section.
 \end{defn}
@@
 \begin{defn}[$\tt{lreplace}$]
-We consider several choices in the previous section.
+We consider several choices in the string operations section.
 \end{defn}
@@
 \begin{proof}
-Correctness implies closure.
+This result is proven for various formulations in the next section.
 \end{proof}
@@
 
-In the S-T-Concat case, note that \ref{thm:total} implies that if $r_1$ and $r_2$ are regular expressions then so is $r_1 \cdot r_2$.
+In the S-T-Concat case, note that Lemma~\ref{thm:total} implies that if $r_1$ and $r_2$ are regular expressions then so is $r_1 \cdot r_2$.
 
@@
 
-To prove type safety and the security theorem for the big step semantics, we
-first prove type safety for a small step semantics in Figure~\ref{sdlambdas}
-and then extend this to the big step semantics in Figure~\ref{dlambdas} by
+To prove type safety and the security theorems for the big step semantics, we
+first prove type safety for a small step semantics in Figure~\ref{fig:sdlambdas}
+and then extend this to the big step semantics in Figure~\ref{fig:dlambdas} by
 proving a correspondence between the semantics.
@@
 
-Note that small step evaluation for $\lambdas$ is terminating:
+\begin{trconj}
+if $\emptyset \vdash e : \sigma$ then $e \mapsto^* v$ such that $v \ \val$.
+\end{trconj}
 %\begin{trthm}[Termination of small step semantics]
-if $\emptyset \vdash e : \sigma$ then $e \mapsto^* v$ such that $v \ \val$.
 %\end{trthm}
@@
 \begin{trlem}[Progress of small step semantics.]
-If $\vdash e : \sigma$ either $e \ \val$ or $e \mapsto e'$ for some $e'$.
+If $\emptyset\vdash e : \sigma$ either $e \ \val$ or $e \mapsto e'$ for some $e'$.
 \end{trlem}
 \begin{proof}
-The proof proceeds by induction on the derivation of $\vdash e : \sigma$.
+The proof proceeds by induction on the derivation of $\emptyset\vdash e : \sigma$.
 \begin{itemize}[label=$ $,itemsep=1ex]
@@
 
-\item \textbf{S-T-Stringin-I}. Suppose $\vdash \rstr{s} : \stringin{s}$.
+\item \textbf{S-T-Stringin-I}. Suppose $\emptyset\vdash \rstr{s} : \stringin{s}$.
 The $\rstr{s} \ \val$ by SS-E-RStr.
 
-\item \textbf{S-T-Concat}. Suppose $\vdash \rconcat{e_1}{e_2} : \stringin{s}$.
+\item \textbf{S-T-Concat}. Suppose $\emptyset\vdash \rconcat{e_1}{e_2} : \stringin{s}$.
 By inversion and induction, $e_1 \mapsto e_1'$ or $e_1 \ \val$ and similarly
@@
 In the remaining case, $e_1 \ \val$ and $e_2 \ \val$.
-But then it follows by canonical forms that $e_1 = \rstr{s_1}$
+But then it follows by Canonical Forms that $e_1 = \rstr{s_1}$
 and $e_2 = \rstr{s_2}$.
-Note that $\rconcat{\rstr{s_1}}{\rstr{s_2}} \mapsto \rstr{s_1s_2}$ by SS-E-Concat.
+Finally, by SS-E-Concat, \\ $\rconcat{\rstr{s_1}}{\rstr{s_2}} \mapsto \rstr{s_1s_2}$.
 
@@
 Suppose $e = \strcase{e_1}{e_2}{x,y.e_3}$. 
-By inversion, $e_1 : \stringin{r}$.
+By inversion,\\ $\emptyset\vdash e_1 : \stringin{r}$.
 %\begin{align}
@@
 %In the former case, $e \mapsto \strcase{e_1}{e_2'}{x,y.e_3}$.
-From this fact, induction, and canonical it follows that $e_1 \mapsto e_1'$ or $e_1 = \rstr{s}$.
+From this fact, induction, and Canonical Forms it follows that $e_1 \mapsto e_1'$ or $e_1 = \rstr{s}$.
 In the former case, $e$ steps by S-E-Case-Left.
@@
 \item \textbf{S-T-Replace}.
-Suppose $e = \rreplace{r}{e_1}{e_2}$ and $e : \stringin{r'}$ where, by inversion of S-T-Replace,
+Suppose $e = \rreplace{r}{e_1}{e_2}$ and $\emptyset\vdash e : \stringin{r'}$ where, by inversion of S-T-Replace,
 \begin{align}
-&\vdash e_1 : \stringin{r_1} \\
-&\vdash e_2 : \stringin{r_2} \\
+&\emptyset\vdash e_1 : \stringin{r_1} \\
+&\emptyset\vdash e_2 : \stringin{r_2} \\
 &\lreplace{r}{r_1}{r_2} = r'
@@
 The only remaining case is where $e_1 \ \val$ and also $e_2 \ \val$.
-But then by canonical forms, $e_1 = \strin{s_1}$ and $e_2 = \strin{s_2}$.
-In this case, $e \mapsto \rstr{\subst{r}}{s_1}{s_2}$ by SS-E-Replace.
+But then by Canonical Forms, $e_1 = \strin{s_1}$ and $e_2 = \strin{s_2}$.
+Therefore, $e \mapsto \rstr{\lsubst{r}{s_1}{s_2}}$ by SS-E-Replace.
 
 \item \textbf{S-T-SafeCoerce}.
-Suppose that $\vdash \rcoerce{r}{e_1} : \stringin{r}$.
+Suppose that $\emptyset\vdash \rcoerce{r}{e_1} : \stringin{r}$.
 By inversion of S-T-SafeCoerce,
 %\begin{align}
-$\vdash e_1 : \stringin{r'}$% \\
+$\emptyset\vdash e_1 : \stringin{r'}$ for $\lang{r'} \subseteq \lang{r}$.% \\
 %&\lang{r'} \subseteq \lang{r}
@@
 If $e_1 \mapsto e_1'$ then $e$ steps by SS-E-SafeCoerce-Step.
-Otherwise, $e_1 \ \val$ and by canonical forms $e_1 = \rstr{s}$.
+Otherwise, $e_1 \ \val$ and by Canonical Forms $e_1 = \rstr{s}$.
 In this case, $e = \rcoerce{r}{\rstr{s}} \mapsto \rstr{s}$ by SS-E-SafeCoerce.
@@
 \item \textbf{S-T-SafeCheck}
-Suppose that $\vdash \rcheck{r}{e_1} : \stringin{r}$.
+Suppose that $\emptyset\vdash \rcheck{r}{e_0}{x.e_1}{e_2} : \stringin{r}$.
 By inversion of S-T-Check:
@@
 In the former case $e$ steps by SS-E-Check-StepRight.
-Otherwise, by canonical forms $e_0 = \rstr{s}$.
-By Lemma~\ref{thm:regexprops}, $s \in \lang{r_0}$ or else not.
-In either case, the SS-E-Check-Ok and SS-E-Check-NotOk steps $e$.
+Otherwise, $e_0 = \rstr{s}$ by Canonical Forms.
+By Lemma~\ref{thm:regexprops}, either $s \in \lang{r_0}$ or $s \not\in\lang{r_0}$.
+In the former case $e$ takes a step by SS-E-Check-Ok.
+In the latter case $e$ takes a step by SS-E-Check-NotOk.
 
@@
 \item \textbf{SS-E-Concat-Left}.
-Suppose $e = \rconcat{e_1}{e_1} \mapsto \rconcat{e_1'}{e_2}$.
-By inversion of S-T-Concat, $\vdash e_1 : \stringin{r_1}$ where $\vdash e : \stringin{r_1r_2}$.
-By induction if $e_1 \mapsto e_1'$ then $\vdash e_1' : \stringin{r_1}$.
-Therefore, $\vdash \rconcat{e_1'}{e_2} : \stringin{r_1r_2}$.
-
-\item {\textbf{SS-E-Concat-Right}}. Similar to SS-E-Concat-Left.
-
-\item \textbf{SS-E-Concat}. Suppose $\rconcat{\rstr{s_1}}{\rstr{s_2}} : \stringin{r_1r_2}$ and
+Suppose $e = \rconcat{e_1}{e_2} \mapsto \rconcat{e_1'}{e_2}$ and $e_1 \mapsto e_1'$.
+By inversion of S-T-Concat, $\emptyset\vdash e_1 : \stringin{r_1}$ where \\$\emptyset\vdash e : \stringin{r_1r_2}$.
+By induction, if $e_1 \mapsto e_1'$ then $\emptyset\vdash e_1' : \stringin{r_1}$.
+Therefore, $\emptyset\vdash \rconcat{e_1'}{e_2} : \stringin{r_1r_2}$.
+
+\item {\textbf{SS-E-Concat-Right}}. \todo{Similar to SS-E-Concat-Left.}
+
+\item \textbf{SS-E-Concat}. Suppose $\emptyset\vdash\rconcat{\rstr{s_1}}{\rstr{s_2}} : \stringin{r_1r_2}$ and
 $\rconcat{\rstr{s_1}}{\rstr{s_2}} \mapsto \rstr{s_1s_2}$.
-Then by inversion $\rstr{s_1} : \stringin{r_1}$ and similarly for $\rstr{s_2}$.
-Therefore, $s_1 \in \lang{r_1}$ and $s_2 \in  \lang{r_2}$ from which it follows by Lemma~\ref{thm:regexprops}
-that $s_1s_2 \in \lang{r_1r_2}$. Therefore, $\vdash \rstr{s_1s_2} : \stringin{r_1r_2}$ by S-T-Rstr.
+Then by inversion $\emptyset\vdash\rstr{s_1} : \stringin{r_1}$ and similarly for $\rstr{s_2}$.
+Therefore, $s_1 \in \lang{r_1}$ and\\ $s_2 \in  \lang{r_2}$ from which it follows by Lemma~\ref{thm:regexprops}
+that $s_1s_2 \in \lang{r_1r_2}$. Therefore, $\emptyset\vdash \rstr{s_1s_2} : \stringin{r_1r_2}$ by S-T-Rstr.
 
 \item \textbf{S-E-Case-Left}.
-Suppose that $e = \strcase{e_1}{e_2}{x,y.e_3} \mapsto \strcase{e_1'}{e_2}{x,y.e_3}$
+Suppose that $e = \strcase{e_1}{e_2}{x,y.e_3}$ and also that
+$e \mapsto \strcase{e_1'}{e_2}{x,y.e_3}$
 and $\emptyset \vdash e : \stringin{r}$.
@@
 \begin{align}
-&\vdash e_1 : \stringin{r} \label{align7} \\
-&\vdash e_2 : \sigma \label{align8} \\ 
+\emptyset&\vdash e_1 : \stringin{r} \label{align7} \\
+\emptyset&\vdash e_2 : \sigma \label{align8} \\ 
 x:\stringin{\lhead{r}}, y : \stringin{\ltail{r}} &\vdash e_3 : \sigma \label{align9}
 \end{align}
-By (\ref{align7}) and the assumption that $e_1 \mapsto e_1'$, it follows by induction that $e_1' : \stringin{r}$.
+By (\ref{align7}) and the assumption that $e_1 \mapsto e_1'$, it follows by induction that $\emptyset\vdash e_1' : \stringin{r}$.
 This fact together with (\ref{align8}) and (\ref{align9}) implies by S-T-Case that
-$\strcase{e_1'}{e_2}{x,y.e_3} : \sigma$.
+$\emptyset\vdash\strcase{e_1'}{e_2}{x,y.e_3} : \sigma$.
 
 \item \textbf{SS-E-Case-Right}. 
-Suppose that $e = \strcase{e_1}{e_2}{x,y.e_3} \mapsto \strcase{e_1}{e_2'}{x,y.e_3}$
+We have that $e = \strcase{e_1}{e_2}{x,y.e_3}$, Suppose $e \mapsto \strcase{e_1}{e_2'}{x,y.e_3}$
 and $\emptyset \vdash e : \stringin{r}$.
@@
 \begin{align}
-&\vdash e_1 : \stringin{r} \label{align10}\\
-&\vdash e_2 : \sigma \label{align11} \\ 
+\emptyset&\vdash e_1 : \stringin{r} \label{align10}\\
+\emptyset&\vdash e_2 : \sigma \label{align11} \\ 
 x:\stringin{\lhead{r}}, y : \stringin{\ltail{r}} &\vdash e_3 : \sigma \label{align12}
 \end{align}
-By (\ref{align11}) and the assumption that $e_2 \mapsto e_2'$, it follows by induction that $e_2' : \stringin{r}$.
+By (\ref{align11}) and the assumption that $e_2 \mapsto e_2'$, it follows by induction that $\emptyset\vdash e_2' : \stringin{r}$.
 This fact together with (\ref{align10}) and (\ref{align12}) implies by S-T-Case that 
-$\strcase{e_1}{e_2'}{x,y.e_3} : \sigma$.
+$\emptyset\vdash\strcase{e_1}{e_2'}{x,y.e_3} : \sigma$.
 
 \item \textbf{SS-E-Case-$\epsilon$-Val}.
-Suppose $e = \strcase{-}{e_2}{-}: \sigma$ and $e \mapsto e_2$.
+Suppose: \begin{align*}
+&e = \strcase{-}{e_2}{-} \\
+&\emptyset\vdash e: \sigma \\
+&e \mapsto e_2
+\end{align*}
 By inversion of S-T-Case, $e_2 : \sigma$.
@@
 Suppose that $e = \strcase{\rstr{as}}{e_2}{x,y.e_3} \mapsto [\rstr{a}, \rstr{s} / x,y] e_3$
-and that $e : \sigma$.
+and that $\emptyset\vdash e : \sigma$.
 By inversion of S-T-Case:
 \begin{align}
-&\vdash \rstr{as} : \stringin{r} \label{align13}\\
-&\vdash \rstr{e_2} : \sigma \label{align14} \\ 
+\emptyset&\vdash \rstr{as} : \stringin{r} \label{align13}\\
+\emptyset&\vdash \rstr{e_2} : \sigma \label{align14} \\ 
 x:\stringin{\lhead{r}}, y : \stringin{\ltail{r}} &\vdash e_3 : \sigma \label{align15}
@@
 We know that $as \in \lang{r}$ by (\ref{align13}) and inversion of S-T-Rstr.
-Therefore, $a \in \lhead{r}$ by definition of ${\sf lhead}$.
+Therefore, $a \in \lang{\lhead{r}}$ by definition of ${\sf lhead}$.
 Furthermore, $\ltail{r} = ... | \delta_a{r} | ...$ by definition of ${\sf ltail}$.
 Note that $s \in \lang{\delta_a{r}}$ by definition of the derivative, and so
-$s \in \ltail{r}$
-
-From these facts about $a$ and $s$ we know by S-T-Rstr that $\vdash \rstr{a} : \stringin{\lhead{r}}$
-and $\vdash \rstr{s} : \stringin{\lhead{r}}$.
-It follows by (\ref{align15}) that  $\vdash [\rstr{a}, \rstr{s} / x,y] e_3 : \sigma$.
+$s \in \lang{\ltail{r}}$
+
+From these facts about $a$ and $s$ we know by S-T-Rstr that $\emptyset\vdash \rstr{a} : \stringin{\lhead{r}}$
+and $\emptyset\vdash \rstr{s} : \stringin{\lhead{r}}$.
+It follows by (\ref{align15}) that  $\emptyset\vdash [\rstr{a}, \rstr{s} / x,y] e_3 : \sigma$.
 
@@
 Suppose that $\rcoerce{r}{s_1} \mapsto \rstr{s_1}$ and that $\emptyset\vdash \rcoerce{r}{s_1} : \stringin{r}$.
-By inversion of S-T-SafeCoerce we know that $s \in \lang{r}$ from which it follows by S-T-Rstr that $\emptyset\vdash s : \stringin{r}$.
+By inversion of S-T-SafeCoerce we know that $s \in \lang{r}$. Therefore, $\emptyset\vdash s : \stringin{r}$.
 
 \item \textbf{Case SS-E-Check-Ok}.
-Suppose that $\rcheck{r}{\rstr{s}}{x.e_1}{e_2} \mapsto \subst{\rstr{s}}{x}{e_1}$,
-$s \in \lang{r}$ and that $\emptyset \vdash \rcheck{r}{\rstr{s}}{x.e_1}{e_2} : \sigma$.
+Suppose $\rcheck{r}{\rstr{s}}{x.e_1}{e_2} \mapsto \subst{\rstr{s}}{x}{e_1}$,
+$s \in \lang{r}$, and $\emptyset \vdash \rcheck{r}{\rstr{s}}{x.e_1}{e_2} : \sigma$.
 By inversion of S-T-Check, $x : \stringin{r} \vdash e_1 : \sigma$.
 Note that $s \in \lang{r}$ implies that $s : \stringin{r}$ by S-T-RStr.
-Therefore, $\emptyset \vdash \subst{\rstr{s}}{x}{e_1} : \sigma$
+Therefore, $\emptyset \vdash \subst{\rstr{s}}{x}{e_1} : \sigma$.
 
 \item \textbf{Case SS-E-Check-NotOk}.
-Suppose that $\rcheck{r}{\rstr{s}}{x.e_1}{e_2} \mapsto e_2$,
-$s \not \in \lang{r}$ and that $\emptyset \vdash \rcheck{r}{\rstr{s}}{x.e_1}{e_2} : \sigma$.
+Suppose $\rcheck{r}{\rstr{s}}{x.e_1}{e_2} \mapsto e_2$,
+$s \not \in \lang{r}$, and $\emptyset \vdash \rcheck{r}{\rstr{s}}{x.e_1}{e_2} : \sigma$.
 By inversion of S-T-Check, $\emptyset \vdash e_2 : \sigma$.
@@
 \begin{proof}
-Follows directly from progress and preservation.
+\todo{Follows directly from progress and preservation.}
 \end{proof}
@@
 
-Before extending these the type safety result fir the small step semantics to the big step semantics, we first establish a correspondence between
+Before extending the previous theorem to the big step semantics, we first establish a correspondence between
 the big step semantics in Figure~\ref{fig:sdlambdas} and the small step semantics in Figure~\ref{fig:dlambdas}.
-We prove the relevant theorems for the $\mapsto$ relation because these proofs are easier. We then extend this result via the correspondence theorem
-to the more concise big step semantics presented in \cite{fulton2014}.
 
@@
 
-From these two facts we know via TR-Step that $\rreplace{r}{e_1}{e_2} \mapsto^* \rreplace{r}{e_1}{e_2}$.
+From these two facts we know via RT-Step that $\rreplace{r}{e_1}{e_2} \mapsto^* \rreplace{r}{e_1}{e_2}$.
 Recall that $\lsubst{r}{s_1}{s_2} = s$, from which the conclusion follows.
@@
 
-Establishing the other direction requires a minor lemma about the relationship between values and reflexivity in the big step semantics.
-
-\begin{trthm}[Small Step Values are Reflexive in the Big Step Semantics]
-If $v \ \val$ in the small step semantics then $v \sreduces v$.
-\end{trthm}
-\begin{proof}
-If $v \ \val$ is derived from SS-E-RStr then S-E-RStr establishes the result.
-Otherwise $v \ \val$ is derived from SS-E-Abs and so S-E-Abs establishes the result.
-\end{proof}
+%Establishing the other direction requires a minor lemma about the relationship between values and reflexivity in the big step semantics.
+%
+%\begin{trthm}[Small Step Values are Reflexive in the Big Step Semantics]
+%If $v \ \val$ in the small step semantics then $v \sreduces v$.
+%\end{trthm}
+%\begin{proof}
+%If $v \ \val$ is derived from SS-E-RStr then S-E-RStr establishes the result.
+%Otherwise $v \ \val$ is derived from SS-E-Abs and so S-E-Abs establishes the result.
+%\end{proof}
 
 \begin{trthm}[Semantic Correspondence for $\lambdas$ (Part II)]
-If $e \vdash \tau$, $e \mapsto^* v$ and $v \ \val$ then $e \sreduces v$.  
+If $\emptyset\vdash e : \sigma$, $e \mapsto^* v$ and $v \ \val$ then $e \sreduces v$.  
 \end{trthm}
@@
 \item \textbf{Case $e = \concat{e_1}{e_2}$}.
-By inversion, $\vdash e_1 : \stringin{r_1}$.
-By Type Safety, canonical forms and termination it follows that $e_1 \mapsto^* \rstr{s_1}$ for some $s_1$.
+By inversion, $\emptyset\vdash e_1 : \stringin{r_1}$.
+By Type Safety, Canonical Forms and Termination it follows that $e_1 \mapsto^* \rstr{s_1}$ for some $s_1$.
 By induction, $e_1 \sreduces \rstr{s_1}$.
@@
 
-Note that $\concat{e_1}{e_2} \mapsto^* \concat{\rstr{s_1}}{\rstr{s_2}} \mapsto \rstr{s_1s_2}$ by SS-E-Concat-LR* and S-E-Concat,
-and so $e \mapsto^* \rstr{s_1s_2}$ by TR-Step. So it suffices to show that $e \sreduces \rstr{s_1s_2}$.
+Note that $\concat{e_1}{e_2} \mapsto^* \concat{\rstr{s_1}}{\rstr{s_2}} \mapsto \rstr{s_1s_2}$ by SS-E-Concat-LR* and S-E-Concat.
+Therefore, $e \mapsto^* \rstr{s_1s_2}$ by RT-Step. So it suffices to show that $e \sreduces \rstr{s_1s_2}$.
 
@@
 \item \textbf{Case $e = \rreplace{r}{e_1}{e_2}$}.
-By inversion of S-T-Replace, $\vdash e_1 : \stringin{r_1}$ for some $r_1$.
-It follows by type safety, termination and canonical forms that $e_1 \mapsto^* \rstr{s_1}$.
+By inversion of S-T-Replace, $\emptyset\vdash e_1 : \stringin{r_1}$ for some $r_1$.
+It follows by Type Safety, Termination and Canonical Forms that $e_1 \mapsto^* \rstr{s_1}$.
 By induction, $e_1 \sreduces \rstr{s_1}$.
@@
 
-Note that $e \mapsto^* \rreplace{r}{\rstr{s_1}}{\rstr{s_2}} \mapsto \rstr{\lsubst{r}{s_1}{s_2}}$ by SS-Replace-LR* and SS-E-Replace
-and so $e \mapsto^* \rstr{\lsubst{r}{s_1}{s_2}}$ by TR-Step.
+Note that $e \mapsto^* \rreplace{r}{\rstr{s_1}}{\rstr{s_2}} \mapsto \rstr{\lsubst{r}{s_1}{s_2}}$ by SS-Replace-LR* and SS-E-Replace.
+Therefore $e \mapsto^* \rstr{\lsubst{r}{s_1}{s_2}}$ by RT-Step.
 
 It suffices to show $e \sreduces \rstr{\lsubst{r}{s_1}{s_2}}$, which follows by S-E-Replace from the facts that $e_1 \sreduces \rstr{s_1}$ and
-similarly for $e_2 \sreduces \rstr{s_2}$.
+$e_2 \sreduces \rstr{s_2}$.
 
 \item \textbf{Case $e = \strcase{e_1}{e_2}{x.y.e_3}$}.
-By inversion, $\vdash e_1 : \stringin{r}$ and $e_2 : \sigma$.
-By type safety, canonical forms and termination $e_1 \mapsto^* \stringin{s_1}$ and by induction $e_1 \sreduces \stringin{s_1}$.
-Similarly, $e_2 \mapsto^* v_2$ and $\vdash e_2 \sreduces v_2$.
-
-By SS-E-Case-LR*, $\strcase{e_1}{e_2}{x.y.e_1} \mapsto^* \strcase{v_1}{v_2}{x.y.e_2}$.
+By inversion, $\emptyset\vdash e_1 : \stringin{r}$ and $e_2 : \sigma$.
+By Type Safety, Canonical Forms and Termination $e_1 \mapsto^* \stringin{s_1}$ and by induction $e_1 \sreduces \stringin{s_1}$.
+Similarly, $e_2 \mapsto^* v_2$ and $\emptyset\vdash e_2 \sreduces v_2$.
+
+By SS-E-Case-LR*, $\strcase{e_1}{e_2}{x,y.e_3} \mapsto^* \strcase{v_1}{v_2}{x,y.e_3}$.
 
@@
 Therefore, by RT-Step, $e \mapsto^* v_2$.
-Also, $e_1 \mapsto \strin{\epsilon}$ and $e_2 \mapsto v_2$ is enough to establish by S-E-Case-$\epsilon$ that $e \sreduces v_2$.
+Recall $e_1 \sreduces \strin{\epsilon}$ and $e_2 \sreduces v_2$, which is enough to establish by S-E-Case-$\epsilon$ that $e \sreduces v_2$.
 
@@
 \begin{proof}
-If $\emptyset \vdash e : \sigma$ then $e \mapsto^* v$ for $v \ \val$ by termination and type safety for the small step semantics.
+If $\emptyset \vdash e : \sigma$ then $e \mapsto^* v$ for $v \ \val$ by termination.
 Therefore, $e \sreduces v$ by part 2 of the semantic correspondence theorem.
@@
 \begin{proof}
-If $\emptyset \vdash e : \stringin{r}$ then $\emptyset \vdash \rstr{s} : \stringin{r}$ by type safety.
+If $\emptyset \vdash e : \stringin{r}$ and $e \sreduces \rstr{s}$ then $\emptyset \vdash \rstr{s} : \stringin{r}$ by Type Safety.
 By inversion of S-T-Rstr, $s \in \lang{r}$.
 \end{proof}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+\section{Proofs of Lemmas and Theorems About $\lambdap$}
+
+\section{Proofs and Lemmas and Theorems About Translation}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+\section{String Substitution and Language Replacement}
+
+\subsection{The Trivial Definition}
+
+\subsection{An Automaton Construction}
+
+Insert Automaton stuff...
+
+\subsection{Toward a Precise Definition}
 
@@
 \caption{Small step semantics for $\lambdas$. Extends \ref{fig:lambda}.}
-\label{fig:dslambdas}
+\label{fig:sdlambdas}
 \end{figure}

### 2014-10-31 att-pldi15/att-pldi15.tex
PLDI15 draft update

%\documentclass[12pt]{article}
\documentclass[10pt,preprint]{sigplanconf}
% The following \documentclass options may be useful:
%
% 10pt          To set in 10-point type instead of 9-point.
% 11pt          To set in 11-point type instead of 9-point.
% authoryear    To obtain author/year citation style instead of numeric.
\usepackage{amsmath}
\usepackage{amssymb} 
\usepackage{amsthm}
\usepackage{ stmaryrd }
\usepackage{mathpartir}
\usepackage[usenames,dvipsnames,svgnames,table]{xcolor}

\input{macros-catlam}

\usepackage{times}
\renewcommand{\ttdefault}{txtt}
\usepackage{alltt}
\usepackage{listings}
\lstset{language=ML,
showstringspaces=false,
basicstyle=\ttfamily\footnotesize,
morekeywords={newcase,extends}}

\usepackage{url}
\usepackage{todonotes}
\lefthyphenmin=5
\sloppy

\newcommand{\moutput}{^{\color{gray}+}}
\newcommand{\rulename}[1]{(#1)}
\def \TirNameStyle #1{\small\rulename{#1}}
\renewcommand{\MathparLineskip}{\lineskiplimit=.6\baselineskip\lineskip=.6\baselineskip plus .2\baselineskip}

\newtheorem{theorem}{Theorem}
\newtheorem{lemma}{Lemma}
\newtheorem{corollary}{Corollary}
\newtheorem{definition}{Definition}
\newtheorem{tyconinvariant}{Tycon Invariant}
\newenvironment{proof-sketch}{\noindent{\emph{Proof Sketch.}}}{\qed}
\makeatletter

% Heather-added packages for the fancy table
\usepackage{longtable}
\usepackage{booktabs}
\usepackage{pdflscape}
\usepackage{colortbl}%
\newcommand{\myrowcolour}{\rowcolor[gray]{0.925}}
\usepackage{wasysym}

\renewcommand\topfraction{0.85}
\renewcommand\bottomfraction{0.85}
\renewcommand\textfraction{0.1}
\renewcommand\floatpagefraction{0.85}

\begin{document}

\conferenceinfo{-}{-} 
\copyrightyear{-} 
\copyrightdata{[to be supplied]}

%\titlebanner{{\tt \textcolor{Red}{{\small Under Review -- distribute within CMU only.}}}}        % These are ignored unless
%\preprintfooter{Distribute within CMU only.}   % 'preprint' option specified.

\title{Modularly Composing Typed Language Fragments}

\authorinfo{}{}{}
%\authorinfo{Cyrus Omar \and Jonathan Aldrich}
 %         {Carnegie Mellon University}
  %         {\{comar, aldrich\}@cs.cmu.edu}

\maketitle
\begin{abstract}
Researchers often describe typed programming languages as fragments or simple calculi, leaving to language designers the task of composing these to form complete languages.  %comparable to those available for library-based embeddings packaged using a modern module system. 
This is not a systematic process: metatheoretic results must be established anew for each composition, guided only notionally by metatheorems derived for simpler systems.
As the design space grows, mechanisms that provide stronger modular reasoning principles than this are needed.%, so that important metatheoretic results need only be established locally and, ultimately, so that language extensions can be treated like libraries. 
%,  Sometimes this fails because certain language features do not ``play well'' together, but precisely characterizing which is elusive given these practices.

%Recent work has started to address this problem, e.g. by showing how to modularly reason about desugarings atop a fixed type system. Our focus here is on safely composing type system fragments. We organize fragments around type constructors, as is usual practice when describing type systems, and sidestep the difficulties of composing abstract syntax (i.e. the expression problem)  by delegating control over a small, uniform abstract syntax in a type-directed manner to logic associated with a type constructor.

In this paper, we begin from first principles with a core calculus, @$\lambda$, specified like many full-scale languages: as a bidirectionally typed translation semantics. Only the $\rightarrow$ type constructor (tycon) is built in; all other external tycons (we show ordered record types and constrained string types) are defined by extending a \emph{tycon context}. Each defines the semantics of associated term-level operators using functions written in a static language where types and translations are values. The semantics maintain several strong metatheoretic guarantees, notably \emph{type safety} and \emph{conservativity}: that \emph{tycon  invariants} maintained under a minimal tycon context will be conserved under any further extended context. Interestingly, mechanized proofs are not needed: problems are caught during typechecking by lifting typed compilation techniques into the semantics and enforcing abstraction barriers around tycons using type abstraction, the same technique that underlies reasoning about ML-style modules.
\end{abstract}

%\category{D.3.2}{Programming Languages}{Language Classifications}[Extensible Languages]
%\category{D.3.4}{Programming Languages}{Processors}[Compilers]
%\category{F.3.1}{Logics \& Meanings of Programs}{Specifying and Verifying and Reasoning about Programs}[Specification Techniques]
%\keywords
%extensible languages; module systems; type abstraction; typed compilation; type-level computation
\section{Introduction}\label{intro}
Typed programming languages are most often described as being composed of \emph{fragments}, each contributing to the language's concrete syntax, abstract syntax, static semantics and dynamic semantics. 
In his textbook, Harper organizes fragments around type constructors, describing each in a different chapter \cite{pfpl}. Languages are then identified by a set of type constructors, e.g. $\mathcal{L}\{{\rightharpoonup}\,{\forall}\,{\mu}\,{\keyw{1}}\,{\times}\,{+}\}$ is the language that includes partial function types, polymorphic types, recursive types, nullary and binary product types and binary sum types (its syntax is shown in Figure \ref{syntax-IL}, discussed below).
Other researchers introduce fragments by defining a simple calculus with a ``catch-all'' constant and base type to stand notionally for all other terms and types that may also be included in some future complete language.\todo{find a really good example, or add this to final PSP rev.??}

In contrast, the usual metatheoretic reasoning techniques for programming languages  (e.g., rule induction) operate on complete language specifications. Each {combination} of fragments must formally be treated as its own monolithic language for which  metatheorems must be established anew, guided only informally by those derived for the smaller systems from which it is notionally composed.

%In both cases, it is left as an extrinsic concern to ensure that the metatheory developed separately is actually conserved when fragments defined in these ways are composed to form a complete language.

This is not an everday problem for programmers only because fragments like those mentioned above are ``general purpose'': they make it possible to \emph{isomorphically embed}  many other fragments as ``libraries''. For example, list types need not be built in because they are isomorphic to the type $\iforall{\alpha}{\imu{t}{\iunit + (\alpha \times t))}}$ (e.g. datatypes in ML, which combine these into a single declaration construct). %Languages providing datatypes al\'a ML  are perhaps most directly oriented around  embeddings into this core. % Leaving the core language simple makes it easier to establish its metatheory and verify that tools, like compilers, are implemented correctly.% They also have encouraging connections to logic. %For example, we do not need to define the type constructor $\fvar{list}$ (indexed by a type) as a fragment (though it is possible to do so) because there is a user-defined datatype constructor, $\texttt{list}$, parameterized by a type, that is isomorphic. If we added the $\fvar{list}$ fragment to the language, it would be entirely redundant semantically: every well-typed term of a type constructed by $\fvar{list}$ corresponds to a well-typed term and a polymorphic recursive sum type such th%Although strings and numbers can be embedded as recursive sum types, it is recognized that this is impractical, so these are also usually included as primitives.
%Establishing an isomorphic embedding of a desirable fragment in terms of general-purpose fragments is not always possible, nor are such embeddings always practical.

Nevertheless, situations do arise where it is not possible to  use these fragments to establish an isomorphic embedding that preserves a desirable fragment's  static and dynamic semantics, including  performance bounds specified by a cost semantics. 
Embeddings are also sometimes too \emph{complex}, as measured by the cost of the extralinguistic computations that are needed to map in and out of the embedding and, if these  must be performed mentally by programmers, considering various human factors. %We will discuss specific examples below.
%When an embedding is too complex, abstraction providers have a few options, discussed in Sec. \ref{prior-work}. When an embedding is not possible, however, or when these options are  insufficient, 
%providers are often compelled to introduce these fragments by extending an existing language, thereby forming a new \emph{dialect}. To save effort, they may do so by forking existing artifacts or leverage tools like compiler generators or language frameworks, also reviewed in Sec. \ref{prior-work}. %Indeed, the proliferation of language dialects constructed for this reason might be taken as an evidence that the core language is not as ``general'', when considered comprehensively, as might be hoped. 
Each time a language is extended with a new fragment for one of these reasons, a new \emph{dialect} is born. Within the ML lineage, for example, dialects abound:
%Reynolds, in a remark that recalls the ``Turing tarpit'' of Perlis \cite{Perl82a}, summarizes the issue \cite{Reynolds94anintroduction}: 
%\begin{quote}
%To say that any reasonable function can be expressed by some program is not to say that it can be expressed by the most reasonable program. It is clear that the language requires a novel programming style. Moreover, it is likely that certain important functions cannot be expressed by their most efficient algorithms.
%\end{quote}
%

\begin{enumerate}
\vspace{-5px}
%\compresslist
\item 
\textbf{General Purpose Fragments:} 
A number of variations on product types, for example, have been introduced in dialects: 
$n$-ary tuples, 
labeled tuples, 
records (identified up to reordering), 
records with width and depth subtyping \cite{Cardelli:1984:SMI:1096.1098}, 
records with update operators\footnote{The Haskell wiki notes that ''No, extensible records are not implemented in GHC. The problem is that the record design space is large, and seems to lack local optima. [...] As a result, nothing much happens.'' \cite{GHCFAQ}} \cite{ocaml-manual}, 
records with mutable fields \cite{ocaml-manual}, 
%field delegation \cite{atlang-gpce14} \todo{gpce submission} 
and 
records with ``methods'' (i.e. pure objects \cite{TSLs}). 
 Sum-like types are also exposed in various ways: 
finite datatypes, 
open datatypes \cite{conf/ppdp/LohH06}, 
hierarchically open datatypes \cite{journals/toplas/MillsteinBC04}, 
polymorphic variants \cite{ocaml-manual} and 
ML-style exception types. Combinations of these manifest themselves as class-based object systems \cite{ocaml-manual}.

\item
\textbf{Specialized Fragments:} Fragments that track specialized static invariants to provide stronger correctness or security guarantees, manage unwieldy lower-level abstractions and run-time systems or control cost are also frequently introduced in dialects, e.g. for data parallelism  \cite{chakravarty2007data}, distributed programming \cite{Murphy:2007:TDP:1793574.1793585}, reactive programming \cite{mandel2005reactiveml}, authenticated data structures \cite{Miller:2014:ADS:2535838.2535851}, databases \cite{Ohori:2011:MSM:2034773.2034815},  units of measure \cite{conf/cefp/Kennedy09} and regular string sanitation \cite{sanitation-psp14}. \verb|sprintf| in the OCaml dialect statically distinguishes format strings from strings.% All of these are implemented as dialects of existing languages, presumably because a strong encoding was not feasible.

\item
\textbf{Foreign Fragments:} A safe and natural foreign function interface (FFI) can be valuable (particularly given this proliferation of dialects). This requires enforcing the type system of the foreign language in the calling language. %Using  a FFI that does not do this can lead to safety issues, even when both languages are separately known to be safe. 
%Safe FFIs generally require direct extensions to the language. 
For example, MLj builds in a safe FFI to Java \cite{Benton:1999:IWW:317636.317791}.
\end{enumerate}
\vspace{-5px}

This \emph{dialect-oriented} state of affairs is unsatisfying. %Language designers are burdened with needing to understand the complete metatheory of every fragment they add to their language to make sure it is not incompatible with existing fragments, so languages change rarely. This decreases the impact of many potentially useful innovations designed by fragment providers (e.g. many academic researchers) by making them unavailable to programmers. 
While programmers can choose from dialects supporting, e.g., a principled approach to distributed programming, or one that builds in support for statically reasoning about units of measure, there may not be an available dialect supporting both. Using different dialects separately for different components of a program is untenable: components written in different dialects cannot always interface safely (i.e. a safe FFI, item 3 above, is needed).

These composability questions not arise when a fragment is expressed as an isomorphic embedding (i.e. as a library) because modern \emph{module systems} enforce abstraction barriers that ensure that the isomorphism need only be established in the ``closed world'' of the module. There are no ``link-time'' proof obligations for clients in the ``open world''. For example, a module defining the semantics of sets in ML can hold the representation of sets abstract, ensuring that any invariants maintained by the functions in the module (e.g. uniqueness, if using a list representation) will hold no matter which other modules are separately in use by a client. %Mechanisms that can help decrease the complexity of an embedding without violating abstraction barriers are thus valuable, and we will lead into our work in Sec. \ref{prior-work} by summarizing them.

When library-based embeddings are not possible, as in the examples above, mechanisms are needed that make it possible to define and  reason in a similarly modular manner about  direct extensions to the semantics of a language. Such a mechanism could ultimately be integrated directly into the language, blurring the distinction between fragments and libraries and decreasing the need for dialects.% Importing fragments that introduce new semantics   would be as easy and reasonable as importing a new module is in ML today. %In the limit, the community could rely on modularly mechanized metatheory and compiler correctness results.% rather than requiring heroic efforts from individual research groups that consider an entire language at once. %Recent work has shown progress on modularly introducing new concrete syntax, reviewed in Sec. \ref{desugaring}. Our focus is on the problem of introducing new semantics.

\paragraph{Contributions} In this paper, we take foundational steps towards this goal by  constructing a minimal but powerful core calculus, @$\lambda$ (the ``actively typed'' lambda calculus). %Despite its minimality, it can host a variety of practical semantic extensions like those described above, while maintaining strong metatheoretic guarantees and, crucially, providing modular reasoning properties. %a calculus introduced briefly in recent work on \emph{active type constructors}  \cite{atlang-gpce14}\todo{TR? Arxiv?}, reviewed in Sec. \ref{overview}. 
Its semantics are structured like those of many modern languages, consisting of an \emph{external language} (EL) governed by a {typed translation semantics} targeting a much simpler \emph{internal language} (IL). 
Rather than building in a monolithic set of external type constructors, however, the  semantics are indexed by a \emph{tycon context}. Each tycon defines the semantics of its operators via functions written in a \emph{static language} (SL) where types and translations are values. %, which provides a form of type-level computation. %The authors demonstrate the expressive power of this technique primarily with an implementation and a number of examples of fragments as libraries, and their core calculus .% (unlike systems that treat the type system as a ``bag of rules'', where non-determinism can arise).

We will begin by giving an overview of the organization  and main judgements of @$\lambda$ in Sec. \ref{atlam}, then discuss how types are constructed and introduce our two main examples, one defining labeled product types with a functional update operator, and the other regular string types, based on a recent core calculus style specification \cite{sanitation-psp14}, in Sec. \ref{types}. We describe how tycons define their associated term-level operator constructors (opcons) in Sec. \ref{external-terms}.  We next give the key metatheoretic properties of the calculus in Sec. \ref{metatheory}, including \emph{type safety} and a key modularity result, which we call  \emph{conservativity}: any invariants that can be established about all values of a type under \emph{some} tycon context (i.e. in some  ``closed world'') are conserved in any further extended tycon context (i.e. in the ``open world''). Interestingly, type system providers need not necessarily provide  mechanized proofs to maintain these guarantees. Instead, the approach we take relies on type abstraction in the internal language. As a result, we are able to borrow the same results that underly modular reasoning in simply-typed languages like ML to reason modularly about typed language fragments. We describe  related work in Sec. \ref{prior-work} and conclude in Sec. \ref{discussion}.

\section{Overview of @$\lambda$}\label{atlam}\label{overview}
\paragraph{External Language}
Programmers interface with @$\lambda$ by writing \emph{external terms}, $e$. The syntax of external terms is shown in Figure \ref{syntax-EL}. %We will describe useful syntactic desugarings atop this syntax as we go on and review recent techniques that permit modularly introducing  desugarings like these in Sec. \ref{desugaring}. Our main focus here is on semantic extensions. 
The static and dynamic semantics are given simultaneously as a \emph{bidirectionally typed translation semantics}, i.e. the key judgements take the form:  \[\esynX{e}{\st\moutput}{\iota\moutput} ~~~~~\text{and}~~~~~ \eanaX{e}{\st}{\iota\moutput}\]
\noindent

These are pronounced ``$e$ (synthesizes / analyzes against) type $\sigma$ and has  translation $\iota$ under typing context $\Upsilon$ and tycon context $\Phi$''. Note that our specifications in this paper are intended to be algorithmic: we indicate ``outputs'' when introducing judgement forms by \emph{mode annotations}, $\moutput$; these are not part of the judgement's syntax. In particular, note that the type is an ``output'' in the synthetic judgement, but an ``input'' in  the analytic judgement.

This basic separation of the EL and IL is commonly used for full-scale language specifications, e.g. the Harper-Stone semantics for Standard ML \cite{Harper00atype-theoretic}. The internal language is purposely kept small, e.g. defining only simple products, to simplify metatheoretic reasoning and compilation. The EL then specifies various useful higher-level constructs, e.g. record types, by translation to the IL. In @$\lambda$, the EL builds in only function types. All other external constructs are defined in the \emph{tycon context}, described in the Sec. \ref{types}.

We choose bidirectional typechecking, also sometimes called \emph{local type inference} \cite{Pierce:2000:LTI:345099.345100}, for two main reasons. The first is once again to justify the practicality of our approach: local type inference is increasingly being used in modern languages (e.g. Scala\todo{citation}) because it eliminates the need for type annotations in many situations while remaining decidable in more situations than whole-function type inference and providing what are widely perceived to be higher quality error messages\todo{citation from Dunfield}. Secondly, it will give us a clean way to reuse the generalized introductory form, $\eintro{\st}{\es}$, and its associated desugarings at many types, in a manner that relates to recent mechanisms supporting type-specific syntax extensions \cite{TSLs}. For example, when we define regular string types, we will be able to reuse standard string literal syntax.

Unlike the Harper-Stone semantics, where external and internal terms were governed by a common type system, in @$\lambda$ each external type, $\st$, maps onto an internal type, $\tau$, called the \emph{type translation} of $\st$. This mapping is specified by the  type translation judgement, $\vdash_\Phi \st \leadsto \tau$, which will be described in Sec. \ref{sec:type-translations}. 
For this reason, this specification style may also be compared to specifications for the first stage of a type-directed compiler, e.g. the TIL compiler for Standard ML \cite{tarditi+:til-OLD}, here lifted ``one level up'' into the semantics of the language itself. As we will see, type safety follows from a property analagous to a correctness condition that arises in typed compilers. Modular reasoning will be based on holding the type translation of $\st$ abstract ``outside'' the tycon.

%We will return to how we control the semantics of the single introductory form in the abstract syntax, $\eintro{\st}{\es}$, on the basis of the type it is being analyzed against, sidestepping issues related to the \emph{expression problem} as a result \todo{cite other paper}\cite{wadler1998expression}. The form $\eother{\iota}$ is a technical device used to quantify over all terms that may arise in a ``future'' tycon context, serving an analagous role to the  ``catchall'' base constants commonly found in minimal calculi, and we will return to its semantics later as well.%We will see how we leverage this bidirectionality when we discuss literal forms below. % The judgement is only well-defined for \emph{valid} tycon contexts, written judgementally as $\vdash \Phi$ and \emph{valid} typing contexts, $\vdash_\Phi \Upsilon$, which we also specify below.
%  \begin{figure*}[t!]
% \small
% \hspace{-5px}\input{example-table}
% %\\~\\
% %$\begin{array}{rcl}
% %e\cdot\conclbl{lbl}(\conclbl{lbl}_1=e_1, \cdots, \conclbl{lbl}_n=e_n)  :=  \etarg{(\conclbl{lbl}; \svar{list}n[\klbl]~\conclbl{lbl}_1~\cdots~\conclbl{lbl}_n)}{e}{e_1; \cdots; e_n}
% %\end{array}$
% \caption{An example external term, $e_{ex}$, in concrete syntax (left), desugared to abstract syntax (right), with static terms shown in green in examples (only). The notation ${\small \scolor{\desugar{...}}}$ stands for an embedding of the indicated ${\small \scolor{\conclbl{label}}}$, ${\small \scolor{\concrx{regular expression}}}$, ${\small \scolor{\concstr{string}}}$ or numeral into the SL, with derived kinds  $\klbl$, $\krx$,  $\kstr$ and $\knat$, not shown. The signatures of helper functions, e.g. ${\small \scolor{\svar{nil}}}$ and ${\small \scolor{\svar{list}n}}$, are shown in Figure \ref{helper-sigs}.}
% \label{example}
% \end{figure*}
\begin{figure}[t]
\small
\hspace{-5px}\input{syntax-table-IL}
\caption{Syntax of {$\mathcal{L}\{{\rightharpoonup}\,{\forall}\,{\mu}\,{\iunit}\,{\times}\,{+}\}$}, our internal language (IL). Metavariable $x$ ranges over term variables and $\alpha$ and $t$ both range over type variables.}
\label{syntax-IL}
\vspace{-10px}
\end{figure}

\begin{figure}[t]
\small
\hspace{-5px}\input{syntax-table-EL}
\caption{Syntax of the external language (EL).}\label{syntax-EL}
\end{figure}

\paragraph{Internal Language} 
%Internal terms, $\iota$, together with internal types, $\tau$, form the \emph{typed internal language}. 
@$\lambda$ relies on a typed internal language supporting type abstraction (i.e. universal quantification over types). We use {$\mathcal{L}\{{\rightharpoonup}\,{\forall}\,{\mu}\,{\iunit}\,{\times}\,{+}\}$}, the syntax for which is shown in Figure \ref{syntax-IL}, as representative of any intermediate language for a typed functional language. %Note that the IL has a fixed semantics; changes to the IL induce different dialects of @$\lambda$.

We assume the statics of the IL are specified in the standard way by judgements for  type formation {$\iDelta \vdash \tau$}, typing context formation { $\iDelta \vdash \iGamma$} and type assignment {$\iDelta~\iGamma \vdash \iota : \tau\moutput$}. The typing and type formation contexts obey standard structural properties (i.e. weakening, exchange and contraction). In examples, we will omit leading $\emptyset$, used as the base case for finite mappings, and $\cdot$, used as the base case for finite sequences (e.g. writing $\Gamma_\text{test} := x : \tau$).

The internal dynamics are specified also in a standard way as a structural operational semantics with a stepping judgement {\small $\iota \mapsto \iota\moutput$} and a value judgement {$\iota~\mathtt{val}$}. The multi-step judgement $\iota \mapsto^{*} \iota\moutput$ is the reflexive, transitive closure of the stepping judgement and the evaluation judgement $\iota \Downarrow \iota'$ is defined iff $\iota \mapsto^{*} \iota'$ and $\iota'~\mathtt{val}$. Both the static and dynamic semantics of the IL can be found in any standard textbook covering typed lambda calculi (we directly follow \cite{pfpl}), so we assume familiarity and key metatheoretic properties.

% We also here define a syntax for simultaneous substitutions, of terms for variables, $\gamma$, and of types for polymorphic type variables, $\delta$, and assume standard judgements ensuring that these are valid with respect to a valid context, $\Delta \vdash \gamma : \Gamma$ and $\vdash \delta : \Delta$. We apply these substitutions to terms, types and typing contexts using the syntax $[\gamma]\iota$, $[\delta]\iota$, $[\delta]\tau$ and $[\delta]\Gamma$ (in some prior work, application of a substitution like this is written using a hatted form, e.g. $\hat\gamma(\iota)$; we intend the same semantics but use a notation more consistent with standard substitution, e.g. $[\iota/x]\iota'$). We will omit  leading $\emptyset$ and $\cdot$ in examples; in specifications, the former is used for  metatheoretic finite mappings, the latter for metatheoretic ordered lists.

\paragraph{Static Language}
\begin{figure}[t]
\small
\hspace{-5px}\input{syntax-table-SL}
\caption{Syntax of the static language (SL). Metavariable $\sx$ ranges over static term variables, $\kalpha$ and $\kvar$ over kind variables and $m$ and $n$ over natural numbers.}
\label{syntax-SL}
\end{figure}
The workhorse of @$\lambda$ is the \emph{static language}, which itself forms a typed lambda calculus where 
%External terms are classified by (external) \emph{types}. 
\emph{kinds}, $\kappa$, classify \emph{static terms}, $\sigma$.  The syntax of the SL is given in Figure \ref{syntax-SL}. The portion of the SL covered by the first row of kinds and first four rows of static terms in the syntax forms an entirely standard  functional programming language consisting of total functions, universal quantification over kinds, inductive kinds (constrained by the standard positivity condition to prevent non-termination), and products and sums. The reader can consider these as forming a total subset of ML or a simply-typed subset of Coq. The semantics also directly follow \cite{pfpl}, so we  omit the details here.  Only three new  kinds will be needed: $\kty$ (Sec. \ref{types}), $\kity$ (Sec. \ref{sec:type-translations}) and $\kitm$ (Sec. \ref{sec:introop}).

The kinding judgement takes the form $\sofkX{\st}{\kappa\moutput}$, where $\kDelta$ and $\kGamma$ are analagous to $\iDelta$ and $\iGamma$ and analagous kind and kinding context formation judgements $\kDelta \vdash \kappa$ and $\kDelta \vdash \kGamma$ are  defined. The natural number $n$ is used as a technical device in our semantics to ensure that the forms shown as being indexed by $n$ in the syntax only arise in a controlled manner internally to prevent ``out of bounds'' issues, as we will discuss; they would have no corresponding concrete syntax so $n$ can be assumed $0$ in user-defined terms.

The dynamic semantics of static terms is defined as a structural operational semantics by the stepping judgement $\sstep{\st}{\argEnv}{\st\moutput}$, the value judgement $\sval{\st}{\argEnv}$ and the error judgement $\serr{\st}{\argEnv}$. Here, $\argEnv$ ranges over \emph{argument environments}, which we will  return to when considering opcons in Sec. \ref{external-terms}. The multi-step judgement $\smanystep{\st}{\argEnv}{\st\moutput}$ is the reflexive, transitive closure of the stepping judgement. The normalization judgement $\seval{\st}{\argEnv}{\st'}$ is defined iff $\smanystep{\st}{\argEnv}{\st'}$ and $\sval{\st'}{\argEnv}$. %${\st}{\memD}{\memG}{\aCtx}{\st\moutput}{\memD\moutput}{\memG\moutput}$, where $\memD$, $\memG$ and $\aCtx$ are also technical devices that will be described later; they too can be ignored from the perspective of user code. We write $\st \Downarrow \st'$ iff $\snorm{\st}{\emptyset}{\cdot}{{\rightharpoonup}; \emptyset; \emptyset}{\st'}{\emptyset}{\cdot}$. \emph{Static values} are normalized static terms. Normalization can also raise an error (to indicate a type error in an external term, or a problem in a tycon definition, as we will discuss), indicated by the judgement $\serrX{\st}$. We omit error propagation rules.%We will refer to the relevant rules as we proceed. 
\begin{figure}[t]
\small
$\begin{array}{lrcl}
\textbf{tycons} & c & ::= & \rightharpoonup ~|~ \tc ~|~ \keyw{other}[m]\\
\textbf{tycon contexts} & \Phi & ::= & \cdot ~|~ \Phi, \tcdef{\tc}{\psi}{\theta}\\
\textbf{tycon structures} & \theta & ::= & \tcstruct{\st}{\omega} \\
\textbf{tycon sigs} & \psi & ::= & \tcsig{\kappa}{\chi}\\
\textbf{opcon structures} & \omega & ::= & \tcstructn{\st} ~|~ \tcstructc{\omega}{op}{\st}\\
\textbf{opcon sigs}& \chi & ::= & \introsig{\kappa} ~|~ \opsigS{\theta}{op}{\kappa}\\
\end{array}$
\caption{Syntax of tycons. Metavariables $\tc$ and $\opname{op}$ range over user-defined tycon and opcon names, respectively, and $m$ ranges over natural numbers.}
\label{syntax-TC}
\end{figure}
\begin{figure*}\begin{mathpar}
\small
\inferrule[k-parr]{
    \sofkX{\st}{\kprod{\kty}{\kty}}
}{
    \sofkX{\sty{\rightharpoonup}{\st}}{\kty}
}

\inferrule[k-ty]{
    \tcdef{\tc}{\tcsig{\ktyidx}{\chi}}{\theta} \in \Phi\\
    \sofkX{\st}{\ktyidx}
}{
    \sofkX{\sty{\tc}{\st}}{\kty}
}

\inferrule[k-otherty]{
    \sofkX{\sttyidx}{\keyw{Nat} \times \kity}
}{
    \sofkX{\sty{\keyw{other}[m]}{\sttyidx}}{\kty}
}\vspace{-10px}
\end{mathpar}
\caption{Kinding rules for types, which take the form $\sty{c}{\sttyidx}$ where $c$ is a tycon and $\sttyidx$ is the type index.}
\label{fig:types}
\end{figure*}

\section{Types}\label{types}

External types, or simply \emph{types}, are static values of kind $\kty$. The introductory forms for kind $\kty$ is $\sty{c}{\st}$, where $c$ is a \emph{tycon} and $\st$ is the \emph{type index}. The syntax  for tycons in Fig. \ref{syntax-TC} specifies that $c$ is either the built-in tycon governing partial functions, $\rightharpoonup$, a user-defined tycon name written in small caps, $\tc$, or an ``other'' tycon, $\keyw{other}[m; \kappa]$ indexed by a natural number $m$ (to allow arbitrarily many such tycons). The kinding rules governing the form $\sty{c}{\st}$ are shown in Figure \ref{fig:types}. The dynamics are simple (see supplement): the index is eagerly normalized and errors propagate. We write $\istype{\st}{\Phi}$  iff $\sofkn{\emptyset}{\emptyset}{\Phi}{0}{\st}{\kty}$ and $\svalNA{\st}$.

The rule \rulename{k-parr} specifies that the type index of partial function types must be a pair of types. We thus say that $\rightharpoonup$ has \emph{index kind} $\kty \times \kty$. To recover a conventional syntax, we can introduce a desugaring from $\st_1 \rightharpoonup \st_2$ to $\sty{\rightharpoonup}{(\st_1, \st_2)}$. %Here, we will benefit by treating it uniformly.

All user-defined tycons must be defined in the \emph{tycon context}, $\Phi$, which is simply a list of tycon definitions. Each tycon defines a name, $\tc$, a \emph{tycon structure}, $\theta$, and a \emph{tycon signature}, $\psi$. We will return to the tycon structure below. Tycon signatures have the form $\tcsig{\ktyidx}{\chi}$, where $\ktyidx$ is the tycon's index kind and $\chi$ is the \emph{opcon signature}, which we   discuss in Sec. \ref{external-terms}. The first premise of \rulename{k-ty} extracts the index kind and the second checks the type index against it.

The rule (k-otherty) governs types constructed by an ``other'' tycon. These will serve only as technical devices to stand in for types other than those in a given tycon context in Sec. \ref{metatheory}. The index of such a type must pair a natural number with a type translation of kind $\kity$, discussed in Sec. \ref{sec:type-translations}.

\paragraph{Examples}
As our first example, consider the user-defined tycon $\tcvar{rstr}$. The index kind of $\tcvar{rstr}$ is $\krx$, which classifies static regular expressions and is defined as an inductive sum kind in the usual way. Types constructed by $\tcvar{rstr}$ will classify \emph{regular strings}, which are statically known to be in the regular language specified by the type index  \cite{sanitation-psp14}. For example, $\stx{title} := \sty{\tcvar{rstr}}{\concrx{.+}}$ classifies non-empty strings and $\stx{conf} := \sty{\tcvar{rstr}}{\concrx{[A-Z]+ \digit\digit\digit\digit}}$ classifies conference names.  The type indices are  here written using standard concrete syntax for concision. Previous work has shown how to define type-specific (here, kind-specific) syntax like this composably in libraries \cite{TSLs}. We define a tycon context containing only the definition of $\tcvar{rstr}$, $\Phi_\text{rstr} := \tcdef{\tcvar{rstr}}{\tcsig{\krx}{\chi_\text{rstr}}}{\theta_\text{rstr}}$.

Our second example is the tycon $\tcvar{lprod}$, which will define a variant of labeled product type (labeled products are like record types, but maintain a row  ordering; record types are also definable in a manner discussed in the supplement, but maintaining an ordering simplifies our discussion). We choose the index kind of $\tcvar{lprod}$ to be $\klist{\klbl \times \kty}$, where list kinds are defined as inductive sums in the usual way, and $\klbl$ classifies static representations of row labels. The tycon context containing only $\tcvar{lprod}$'s definition is $\small\Phi_\text{lprod} := \tcdef{\tcvar{lprod}}{\tcsig{\klist{\klbl \times \kty}}{\chi_\text{lprod}}}{\theta_\text{lprod}}$.

In the tycon context containing both tycon definitions, $\Phi_\text{rstr}\Phi_\text{lprod}$, we can define a labeled product type classifying conference papers, $\stx{paper} := \sty{\tcvar{lprod}}{\{\conclbl{title} : \stx{title}, \conclbl{conf} : \stx{conf}\}}$. Note that $\istype{\stx{paper}}{\Phi_\text{rstr}\Phi_\text{lprod}}$ and we again use kind-specific syntax, in this case for $\klbl$ and $\klist{\klbl \times \kty}$.

\subsection{Type Case Analysis}
A type $\st$ can be case analyzed against a known tycon $c$ using $\stycase{c}{\st}{\sx}{\st_1}{\st_2}$. If the value of $\st$ is constructed by $c$, its type index is bound to $\sx$ and the branch $\st_1$ is taken. For totality, a default branch, $\st_2$, must also be provided.  For example, the kinding rule for when $c$ is user-defined is below. 
\begin{mathpar}
\small
\inferrule[k-tycase]{
    \sofkX{\st}{\kty}\\
    \tcdef{\tc}{\tcsig{\ktyidx}{\chi}}{\theta} \in \Phi\\
    \sofk{\kDelta}{\kGamma, \sx :: \ktyidx}{\Phi}{\st_1}{\kappa}\\
    \sofkX{\st_2}{\kappa}
}{
    \sofkX{\stycase{\tc}{\st}{\sx}{\st_1}{\st_2}}{\kappa}
}
\end{mathpar}
% \begin{mathpar}\small
% \inferrule[s-tycase-fail-2]{ }{
%     \sstep{\stycase{c}{\sotherty{m}{\tau}}{x}{\st_1}{\st_2}}{\argEnv}{\st_2}
% }
% \end{mathpar}

The rule for arrow types is analagous, but no rule for tycons of the form $\keyw{other}[m]$ is defined.

Put another way, types can be thought of as arising from a distinguished ``open datatype'' defined by the tycon context\todo{citation}. %The dynamics (see supplement) are again straightforward, consistent with this intuition.% We will see an example of its use in Sec. \ref{sec:targops}.
\begin{figure}\hfill \fbox{$\vdash \Phi$}\vspace{-25px}\begin{mathpar}
\small
\inferrule[tcc-ext]{
    \vdash \Phi\\
    \tc \notin \text{dom}(\Phi)\\\\
    \psi = (\tcsig{\ktyidx}{\chi})\\
    \keq{\emptyset}{\ktyidx}\\\\
    \theta = (\tcstruct{\stx{schema}}{\omega})\\
    \sofkz{\emptyset}{\emptyset}{\Phi}{\stx{schema}}{\ktyidx \rightarrow \kity}\\\\
    \vdash_{\Phi, \tcdef{\tc}{\psi}{\theta}} \omega \sim \psi
}{
    \vdash \Phi, \tcdef{\tc}{\psi}{\theta}
}\vspace{-8px}
\end{mathpar}
\caption{Tycon context well-definedness. We omit the empty case (tcc-emp) for concision.}
\label{fig:tycon-ctxs}
\end{figure}
\subsection{Tycon Context Well-Definedness}
The tycon context well-definedness judgement, $\vdash \Phi$, shown in Figure \ref{fig:tycon-ctxs}, requires that all tycon names are unique and performs additional checks,  described below.

\subsection{Type Equivalence} 
The first check simplifies the handling of type equivalence: type index kinds must be \emph{equality kinds}, i.e. those for which semantic equivalence implies syntactic equality at normal form. We define these by the judgement $\keq{\kDelta}{\kappa}$ (see supplement). 
Equality kinds are similar to equality types as found in Standard ML\todo{citation}. The main implication of this choice is that type indices cannot contain static functions.

\subsection{Type Translations}\label{sec:type-translations}
Each tycon {computes} translations for the types it constructs as a function of each type's index by specifying a \emph{translation schema} as the first component of the {tycon structure}, $\theta$. For a tycon with index kind $\ktyidx$, the translation schema must have kind $\ktyidx \rightarrow \kity$, checked by (tcc-ext).

The kind $\kity$ has a single introductory form, $\sqity{\qity}$, where $\qity$ is a \emph{translational internal type}. Each form in the syntax for internal types, $\tau$,  corresponds to a form in the syntax of translational internal types, $\qity$. For example, our translation schema for $\tcvar{rstr}$  simply chooses to ignore the type index and represent all regular strings internally as strings, of internal type abbreviated $\keyw{str}$. We abbreviate the corresponding translational internal type $\hat{\keyw{str}}$ and define the translation schema as $\small\theta_\text{rstr} := \tcstruct{\slam{\krx}{\svar{tyidx}}{\sqity{\hat{\keyw{str}}}}}{\omega_\text{rstr}}$.

The kinding rules and operational semantics for these shared forms simply proceed recursively, e.g.
\begin{mathpar}
\small
\inferrule[k-ity-prod]{
    \sofkX{\sqity{\qity_1}}{\kity}\\
    \sofkX{\sqity{\qity_2}}{\kity}
}{
    \sofkX{\sqity{\qity_1 \times \qity_2}}{\kity}
}
\end{mathpar}
%The operational semantics for shared forms are also straightforwardly recursive (see supplemental material).

The syntax for translational internal types additionally includes an ``unquote'' form,  $\qtuq{\st}$, so that they can be constructed compositionally, as well as a form, $\srep{\st}$, that allows one type's translation to refer to the translation of another type $\st$. \begin{mathpar}\small
\inferrule[k-ity-unquote]{
    \sofkX{\st}{\kity}   
}{
    \sofkX{\sqity{\qtuq{\st}}}{\kity}
}

\inferrule[k-ity-trans]{
    \sofkX{\st}{\kty}
}{
    \sofkX{\sqity{\srep{\st}}}{\kity}
}
\end{mathpar}
The unquote form is eliminated during normalization, while references to the translation of a type are retained in values of kind $\kity$. The key rules in the dynamics are: 
\begin{mathpar}
\small
\inferrule[s-ity-unquote-elim]{
    \sval{\sqity{\qity}}{\argEnv}
}{
    \sstep{\sqity{\qtuq{\sqity{\qity}}}}{\argEnv}{\sqity{\qity}}
}

\inferrule[s-ity-trans-val]{
    \sval{\st}{\argEnv}
}{
    \sval{\sqity{\srep{\st}}}{\argEnv}
}
\end{mathpar}

We choose a translation schema for $\tcvar{lprod}$ that generates nested binary product types by folding over the type index and referring to the translations of the types therein (though this is not the only workable choice, e.g. we could also have used a list). We assume $\small\svar{listfold} :: \kforall{\kalpha_1}{\kforall{\kalpha_2}{\klist{\kalpha_1}\rightarrow\kalpha_2\rightarrow(\kalpha_1\rightarrow\kalpha_2\rightarrow\kalpha_2)\rightarrow\kalpha_2}}$ in defining $\theta_\text{lprod} := \keyw{trans}=\stx{lprod/trans}$  where $\stx{lprod/trans}:=$
\[\small
\begin{array}{l}
\slam{\klist{\klbl \times \kty}}{\svar{tyidx}}{\skap{\kity}{\skap{\klbl\times\kty}{\svar{listfold}}}~\svar{tyidx}~\sqity{\iunit}~\\
  \quad (\lambda \svar{h}{:}\klbl \times \kty.\lambda \svar{r}{:}\kity.\sqity{\srep{{\ssnd{\svar{h}}}}\times\qtuq{\svar{r}}}}
\end{array}\]
Evaluating this translation schema with the index of $\stx{paper}$, for example, produces the value $\stx{paper/trans} := \sqity{\qity_\text{paper/trans}}$ where $\small\qity_\text{paper/trans} := \srep{\stx{title}}\times(\srep{\stx{conf}}\times \iunit)$. Note that we do not include logic to optimize away the trailing unit type for simplicity (and to again emphasize that many translations are possible for any given type).%For simplicity, and to make the point that the choice of representation has only tycon-local implications, our translation schema does not optimize away

\subsubsection{Selective Type Translation Abstraction}\label{sec:selective-type-translation-abstraction}
References to translations of other types are maintained in values of kind $\kity$ like this to  allow us to selectively hold them abstract, which will be the key to our main results. This can be thought of as analagous to the process in ML by which the true identity of an abstract type in a module is held abstract outside the module until after typechecking. The judgement $\small\tdeabs{\Phi}{c}{\qity}{\memD}{\ity\moutput}{\memD\moutput}$ relates a normalized translational internal type $\qity$ to an internal type $\tau$, called a corresponding \emph{abstracted type translation} because references to translations of types constructed by a tycon other than those constructed by a ``delegated tycon'', $c$, are replaced by universally quantified type variables, $\alpha$. For example, $\small\tdeabs{\Phi_\text{rstr}\Phi_\text{lprod}}{\tcvar{lprod}}{\qity_\text{paper/trans}}{\emptyset}{\tau_\text{paper/abs}}{\memD_\text{paper/abs}}$
where $\small\tau_\text{paper/abs} := \alpha_1 \times (\alpha_2 \times \iunit)$ and 
$\memD_\text{paper/abs}  := \stx{title} \leftrightarrow \keyw{str}/\alpha_1, \stx{conf} \leftrightarrow \keyw{str}/\alpha_2$.

Each unique type has a unique type variable associated with it, tracked by the \emph{type translation store}, $\memD$, which maps from types, $\st$, to their (concrete) translations, $\tau$, and the type variables, $\alpha$, which appear in their place in the abstract type translation: $\memD ::= \emptyset ~|~ \memD, \st \leftrightarrow \ity/\alpha$. 
Each type translation store induces a \emph{type substitution}, $\delta$, and a corresponding internal type formation context, $\Delta$, according to the judgement $\memD \leadsto \delta : \Delta$. Type substitutions simply define an $n$-ary substitution for type variables, $\delta ::= \emptyset ~|~ \delta, \tau/\alpha$. For example, $\memD_\text{paper/abs} \leadsto \delta_\text{paper/abs} : \Delta_\text{paper/abs}$ where $\delta_\text{paper/abs} := \keyw{str}/\alpha_1, \keyw{str}/\alpha_2$ and $\Delta_\text{paper/abs} := \alpha_1, \alpha_2$. 
We can apply type substitutions to internal types, terms and typing contexts, written $[\delta]\ity$, $[\delta]\iota$ and $[\delta]\Gamma$, respectively. For example, $[\delta_\text{paper/abs}]\tau_\text{paper/abs}$ is $\tau_\text{paper} := \keyw{str} \times (\keyw{str} \times \iunit)$, i.e. the (concrete) translation of $\stx{paper}$. Indeed, we can now give the rule defining the type translation judgement, $\vdash_\Phi \st \leadsto \tau$, mentioned in Sec. \ref{overview}. We simply determine any selectively abstract translation, then apply the substitution:
\begin{mathpar}\small
\inferrule[conc-ty-trans]{
    \istype{\st}{\Phi}\\
    \tdeabs{\Phi}{\tc}{\srep{\st}}{\emptyset}{\tau}{\memD}\\
    \memD \leadsto \delta : \Delta
}{
    \vdash_\Phi \st \leadsto [\delta]\tau
}
\end{mathpar}

%This process of selectively holding the translations of all types constructed by any tycon other than the ``delegated tycon'' abstract will be critical for conservativity to hold below. Here, the translation of the type is being held abstract, rather than the type itself, and type translations are computed from indices, rather than declared ``literally''.

The selective type abstraction judgement is defined by recursing generically over sub-terms until terms of the form $\srep{\st}$ are encountered (see supplement).

The translation of partial function types is direct and is not held abstract, so that lambdas can be used as the sole binding construct in the language:
\begin{mathpar}
\small
\inferrule[abs-parr]{
    \tdeabs{\Phi}{c}{\srep{\st_1}}{\memD}{\ity_1}{\memD'}\\
    \tdeabs{\Phi}{c}{\srep{\st_2}}{\memD'}{\ity_2}{\memD''}
}{
    \tdeabs{\Phi}{c}{\srep{\sty{\rightharpoonup}{(\st_1, \st_2)}}}{\memD}{\tau_1 \rightharpoonup \tau_2}{\memD''}
}
\end{mathpar}
The translation of a user-defined types constructed by the delegated tycon is determined by calling the translation schema and checking that the type translation it generates refers only to type variables generated from $\memD'$:
\begin{mathpar}
\small
\inferrule[abs-tc-local]{
    \tcdef{\tc}{\psi}{\tcstruct{\stx{schema}}{\omega}} \in \Phi\\\\
    \sevalNA{\stx{schema}(\sttyidx)}{\sqity{\qity}}\\
    \tdeabs{\Phi}{\tc}{\qity}{\memD}{\ity}{\memD'}\\\\
    \memD' \leadsto \delta : \Delta\\
    \Delta \vdash \tau
}{
    \tdeabs{\Phi}{\tc}{\srep{\sty{\tc}{\sttyidx}}}{\memD}{\tau}{\memD'}
}
\end{mathpar}
The translation of a user-defined type constructed by any tycon other than the delegated tycon is added to the store (the supplement has the simple rule for retriving it once there):
\begin{mathpar}
\small
\inferrule[abs-tc-foreign-new]{
    c \neq \tc\\
    \sty{\tc}{\sttyidx} \notin \text{dom}(\memD)\\
    \tcdef{\tc}{\psi}{\tcstruct{\stx{schema}}{\omega}} \in \Phi\\\\
    \sevalNA{\stx{schema}~\sttyidx}{\sqity{\qity}}\\
    \tdeabs{\Phi}{\tc}{\qity}{\memD}{\ity}{\memD'}\\\\
    \memD' \leadsto \delta : \Delta\\
    \Delta \vdash \tau\\
    (\alpha~\text{fresh})
}{
    \tdeabs{\Phi}{c}{\srep{\sty{\tc}{\sttyidx}}}{\memD}{\alpha}{\memD', \sty{\tc}{\sttyidx} \leftrightarrow \tau/\alpha}
}
\end{mathpar}
The translation of an ``other'' type is given directly in its index (the rule where it is held abstract is in the supplement):
\begin{mathpar}
\small
\inferrule[abs-tc-other-new]{
    \tdeabs{\Phi}{\keyw{other}[m]}{\qity}{\memD}{\tau}{\memD'}
}{
    \tdeabs{\Phi}{\keyw{other}[m]}{\srep{\sty{\keyw{other}[m]}{(\stx{nat}, \sqity{\qity}}}}{\memD}{\tau}{\memD'}
}
\end{mathpar}

\subsection{Typing Contexts}
We must also define a translation judgement for external typing contexts $\vdash_\Phi \Upsilon \leadsto \Gamma$ that checks that $\Upsilon$ contains valid types, then generates the corresponding internal typing context $\Gamma$ from their translations (see supplement).
%\vspace{-5px}
 %We have that external typing contexts obey the standard structural congruences: weakening, exchange and contraction.
\section{Typing and Translation of External Terms}\label{external-terms}
Having established how types are constructed, and how they determine abstract and from there concrete translations to internal types, we can finally give the typing and translation rules for external terms, shown in Figure \ref{typing}.

Because we are defining a bidirectional type system, a subsumption rule is needed to allow synthetic terms to be analyzed against an equivalent type. Per above, equivalent types must be  syntactically identical at normal form, and we consider analysis only if $\istype{\st}{\Phi}$, so the rule (subsume) is straightforward.

To use an analytic term in a synthetic position, the programmer must provide a type ascription, written $e : \st$. The ascription is kind checked and normalized to a type before being used for analysis by rule (ascribe).

Variables and functions behave in the standard manner given our definitions of types and type translations (used to generate the ascription on the lambda in the IL, which is intrinsically rather than bidirectionally typed). We use Plotkin's fixpoint operator for general recursion (cf. \cite{pfpl}), and define both only in analytic positions for simplicity.

\begin{figure*}[t]
\small\fbox{$\eanaX{e}{\st}{\iota}$}~
\fbox{$\esynX{e}{\st}{\iota}$}
\begin{mathpar}
\inferrule[subsume]{
    \esynX{e}{\st}{\iota}
}{
    \eanaX{e}{\st}{\iota}
}

\inferrule[ascribe]{
  \sofkz{\emptyset}{\emptyset}{\Phi}{\st}{\kty}\\
  \snormT{\st}{\st'}\\\\
  \eanaX{e}{\st'}{\iota}
}{
  \esynX{\easc{e}{\st}}{\st'}{\iota}
}

\small\inferrule[syn-var]{
  x \Rightarrow \st \in \Upsilon
}{
  \esynX{x}{\st}{x}
}

\inferrule[ana-fix]{
  \eana{\Upsilon, x \Rightarrow \st}{\Phi}{e}{\st}{\iota}\\\\
  \sfinalrepX{\st}{\tau}
}{
  \eanaX{\efix{x}{e}}{\st}{\ifix{\tau}{x}{\iota}}
}

\inferrule[ana-lam]{
    \eana{\Upsilon, x \Rightarrow \st_1}{\Phi}{e}{\st_2}{\iota}\\\\
    \sfinalrepX{\st_1}{\tau_1}
}{
    \eanaX{\eanalam{x}{e}}{\sty{\rightharpoonup}{(\st_1, \st_2)}}{\ilam{\tau_1}{x}{\iota}}
}

\inferrule[syn-ap]{
  \esynX{e_1}{\sty{\rightharpoonup}{(\st_1, \st_2)}}{\iota_1}\\
  \eanaX{e_2}{\st_2}{\iota_2}
}{
  \esynX{\eap{e_1}{e_2}}{\st_2}{\iap{\iota_1}{\iota_2}}
}

\inferrule[ana-intro]{
  \tcdef{\tc}{\tcsig{\_}{\chi}}{\tcstruct{\_}{\omega}} \in \Phi\\\\
  \introsig{\klitidx} \in \chi\\
  \sofkz{\emptyset}{\emptyset}{\Phi}{\stmidx}{\klitidx}\\\\
  \keyw{ana~intro}={\stx{def}} \in \omega\\
  |\es| = n\\
    \keyw{args}(n)=\stx{args}\\\\
  \stx{def}~\sttyidx~\stmidx~\stx{args} \Downarrow_{\es;\Upsilon;\Phi} \sqitm{\qitm}\\\\
  %\snorm{\stx{def}~\sttyidx~\stmidx~\stx{args}}{\emptyset}{\memG_0}{{\rightharpoonup}, \tc; \Upsilon; \Phi}{\sqitm{\itm_\text{abs}}}{\memD}{\memG}\\\\
  %\snorm{\srep{\sty{\tc}{\sttyidx}}}{\memD}{\cdot}{{\rightharpoonup}, \tc;\emptyset;\Phi}{\sqity{\ity_\text{abs}}}{\memD'}{\cdot}\\\\
  \edeabs{\tc}{{\es;\Upsilon;\Phi}}{\qitm}{\emptyset}{\emptyset}{\iota_\text{abs}}{\memD}{\memG}\\
  \memG \leadsto \gamma : \Gamma\\\\
  \tdeabs{\Phi}{\tc}{\srep{\sty{\tc}{\sttyidx}}}{\memD}{\ity_\text{abs}}{\memD'}\\
  \memD' \leadsto \delta : \Delta\\\\
  \Delta~\Gamma \vdash \iota_\text{abs} : \tau_\text{abs}
}{
  \eanaX{\eintro{\stmidx}{\es}}{\sty{\tc}{\sttyidx}}{[{\delta}][{\gamma}]\iota}
}

\inferrule[syn-targ]{
  \esynX{e_\text{targ}}{\sty{\tc}{\sttyidx}}{\iota_\text{targ}}\\\\
  \tcdef{\tc}{\tcsig{\_}{\chi}}{\tcstruct{\_}{\omega}} \in \Phi\\\\
  \opsig{\opname{op}}{\klitidx} \in \chi\\
  \sofkz{\emptyset}{\emptyset}{\Phi}{\stmidx}{\klitidx}\\\\
  \keyw{syn~}\opname{op}={\stx{def}} \in \omega\\
  |e_\text{targ}; \es| = n\\
    \keyw{args}(n)=\stx{args}\\\\
  \stx{def}~\sttyidx~\stmidx~\stx{args} \Downarrow_{(e_\text{targ}; \es);\Upsilon;\Phi} (\st, \sqitm{\qitm})\\\\
  %\snorm{\stx{def}~\sttyidx~\stmidx~\stx{args}}{\emptyset}{\memG_0}{{\rightharpoonup}, \tc; \Upsilon; \Phi}{\sqitm{\itm_\text{abs}}}{\memD}{\memG}\\\\
  %\snorm{\srep{\sty{\tc}{\sttyidx}}}{\memD}{\cdot}{{\rightharpoonup}, \tc;\emptyset;\Phi}{\sqity{\ity_\text{abs}}}{\memD'}{\cdot}\\\\
  \edeabs{\tc}{{(e_\text{targ}; \es);\Upsilon;\Phi}}{\qitm}{\emptyset}{\emptyset}{\iota_\text{abs}}{\memD}{\memG}\\
  \memG \leadsto \gamma : \Gamma\\\\
  \tdeabs{\Phi}{\tc}{\srep{\st}}{\memD}{\ity_\text{abs}}{\memD'}\\
  \memD' \leadsto \delta : \Delta\\\\
  \Delta~\Gamma \vdash \iota_\text{abs} : \tau_\text{abs}
}{
    \esynX{\etarg{\opname{op}}{\sttmidx}{e_\text{targ}}{\es}}{\st}{[\delta][\gamma]\iota_\text{abs}}
}

\inferrule[ana-intro-other]{
    \sofkz{\emptyset}{\emptyset}{\Phi}{\stx{def}}{\kargs \rightarrow \kitm}\\\\
    |\es| = n\\
    \keyw{args}(n)=\stx{args}\\
    \stx{def}~\stx{args} \Downarrow_{\es; \Upsilon; \Phi} \sqitm{\qitm}\\\\
    \edeabs{\keyw{other}[m]}{\es; \Upsilon;\Phi}{\qitm}{\emptyset}{\emptyset}{\itm_\text{abs}}{\memD}{\memG}\\
    \memG \leadsto \gamma : \Gamma\\\\
    \tdeabs{\Phi}{\keyw{other}[m]}{\srep{\sty{\keyw{other}[m]}{\sttyidx}}}{\memD}{\tau_\text{abs}}{\memD'}\\
    \memD' \leadsto \delta : \Delta\\\\
    \Delta~\Gamma \vdash \iota_\text{abs} : \tau_\text{abs}
}{
    \eanaX{\eintro{\stx{def}}{\es}}{\sty{\keyw{other}[m]}{\sttyidx}}{[\delta][\gamma]\iota_\text{abs}}
}
~~~~~~~~
\inferrule[syn-targ-other]{
\esynX{e_\text{targ}}{\sty{\keyw{other}[m]}{\sttyidx}}{\iota_\text{targ}}\\\\
  \sofkz{\emptyset}{\emptyset}{\Phi}{\stx{def}}{\kargs\rightarrow(\kty\times\kitm)}\\\\
  |e_\text{targ}; \es| = n\\
    \keyw{args}(n)=\stx{args}\\
  \stx{def}~\stx{args} \Downarrow_{(e_\text{targ}; \es);\Upsilon;\Phi} (\st, \sqitm{\qitm})\\\\
  %\snorm{\stx{def}~\sttyidx~\stmidx~\stx{args}}{\emptyset}{\memG_0}{{\rightharpoonup}, \tc; \Upsilon; \Phi}{\sqitm{\itm_\text{abs}}}{\memD}{\memG}\\\\
  %\snorm{\srep{\sty{\tc}{\sttyidx}}}{\memD}{\cdot}{{\rightharpoonup}, \tc;\emptyset;\Phi}{\sqity{\ity_\text{abs}}}{\memD'}{\cdot}\\\\
  \edeabs{\keyw{other}[m]}{{(e_\text{targ}; \es);\Upsilon;\Phi}}{\qitm}{\emptyset}{\emptyset}{\iota_\text{abs}}{\memD}{\memG}\\
  \memG \leadsto \gamma : \Gamma\\\\
  \tdeabs{\Phi}{\keyw{other}[m]}{\srep{\st}}{\memD}{\ity_\text{abs}}{\memD'}\\
  \memD' \leadsto \delta : \Delta\\\\
  \Delta~\Gamma \vdash \iota_\text{abs} : \tau_\text{abs}
}{
     \esynX{\etarg{\opname{op}}{\stx{def}}{e_\text{targ}}{\es}}{\st}{[\delta][\gamma]\iota_\text{abs}}
}
\end{mathpar}
\caption{Typing}
\label{typing}
\end{figure*}
\begin{figure*}[t]
\small\fbox{$\vdash_\Phi \omega \sim \psi$}
\vspace{-25px}
\begin{mathpar}\small
\hspace{50px}\inferrule[ocstruct-intro]{
    \introsig{\klitidx} \in \chi\\
    \emptyset \vdash \klitidx\\\\
    \sofkz{\emptyset}{\emptyset}{\Phi}{\stx{def}}{\ktyidx \rightarrow \klitidx \rightarrow \kargs \rightarrow \kitm}
}{
    \vdash_\Phi \tcstructn{\stx{def}} \sim \tcsig{\ktyidx}{\chi}
}

\inferrule[ocstruct-targ]{
    \vdash_\Phi \omega \sim \tcsig{\ktyidx}{\chi}\\
    \opname{op}\notin\text{dom}(\chi)\\
    \emptyset \vdash \klitidx\\\\
    \sofkn{\emptyset}{\emptyset}{\Phi}{0}{\stx{def}}{\ktyidx \rightarrow \klitidx \rightarrow \kargs \rightarrow (\kty \times \kitm)}
}{
    \vdash_\Phi \tcstructc{\omega}{op}{\stx{def}} \sim \tcsig{\ktyidx}{\chi, \opsig{\opname{op}}{\klitidx}}
}
\end{mathpar}\vspace{-5px}
\caption{Checking opcon structures against tycon signatures.}
\label{ocstruct}
\end{figure*}

\subsection{Generalized Introductory Operations}\label{sec:introop}
The translation of the generalized introductory form, $\eintro{\sttmidx}{\es}$, is determined by the tycon of the type it is being analyzed against as a function of the type's index, the \emph{term index}, $\sttmidx$, and the \emph{argument list}, $\es$.

Before discussing rules (ana-intro) and (ana-intro-other), we note that we can recover a variety of standard concrete introductory forms by a purely syntactic desugaring to this abstract form (and thus allow their use at more than one type). For example, for regular strings we will be able to use the string literal form, $\texttt{"s"}$, which desugars to $\eintro{\texttt{"s"}_\text{SL}}{\cdot}$. The term index is the corresponding static value of kind $\keyw{Str}$, indicated by a subscript for clarity. Similarly, for labeled products, records, objects and so on, we have a generalized labeled collection form, $\{\mathtt{lbl}_1=e_1, \ldots, \mathtt{lbl}_n=e_n\}$, that desugars to $\eintro{\texttt{[}\mathtt{lbl}_1, \ldots, \mathtt{lbl}_n\texttt{]}}{e_1; \ldots; e_n}$, i.e. a list constructed from the row labels is the term index and the corresponding row values are the arguments. In both cases, the term index captures  static portions of the concrete form and the arguments capture all external sub-terms. Additional desugarings are shown in the supplement and a technique based on \cite{TSLs} could be introduced to allow tycon providers to define more such desugarings  composably.

Let us derive $\small\eana{\Upsilon_\text{ex}}{\Phi_\text{rstr}\Phi_\text{lprod}}{e_\text{ex}}{\stx{paper}}{\iota_\text{ex}}$ where we define an example typing context $\small\Upsilon_\text{ex}  := title \Rightarrow \stx{title}$ and $\small e_\text{ex} := \{\conclbl{title}=title, \conclbl{conf}=\texttt{"EXMPL 2015"}\}$. The  translation will be $\small\iota_\text{ex} := (title, (\texttt{"EXMPL 2015"}_\text{IL}, ()))$, where $\small\texttt{"EXMPL 2015"}_\text{IL}$ is an internal string (of internal type $\keyw{str}$).

The first premise of (ana-intro) extracts the tycon definition for the tycon of the type the literal is being analyzed against from $\Phi$. In this example, this is $\tcvar{lprod}$. We will use this as the \emph{delegated tycon} in the final premises of the rule.

The second premise extracts the \emph{intro index kind}, $\klitidx$, from the \emph{opcon signature}, $\chi$, of the tycon signature and the third premise checks the provided term index against this kind. This is simply the kind of term index expected by the tycon, e.g., $\tcvar{lprod}$ specifies $\klist{\klbl}$, so that it can use the labeled collection form, while $\tcvar{rstr}$ specifies an intro index kind of $\keyw{Str}$, so that it can use the string literal form. %\[\small\begin{array}{lcl}
%\chi_\text{lprod} & := & \introsig{\klist{\klbl}}, \chi_\text{lprod/targops}\\
%\chi_\text{rstr} & := & \introsig{\keyw{Str}}, \chi_\text{rstr/targops}
%\end{array}
%\]

The fourth premise extracts the \emph{intro opcon definition} from the \emph{opcon structure}, $\omega$, of the tycon structure, calling it $\stx{def}$. This is a static function that is applied, in the seventh premise, to determine whether the term is well-typed, raising an error if not or determining the translation of the term if so. The function has access to the type index, the term index and an interface to the list of arguments. The  judgement $\vdash_\Phi \omega \sim \psi$, which appeared as the final premise of the rule (tcc-ext),  checks that the intro opcon definition is well-kinded (rule (ocstruct-intro) in Figure \ref{ocstruct}).

For example, the opcon structure of $\tcvar{rstr}$ is  $\small\omega_\text{rstr} := \tcstructn{\stx{rstr/intro}}; \omega_\text{rstr/targops}$ where $\small\stx{rstr/intro}:=$
\[\small\begin{array}{l}
    \slam{\krx}{\svar{tyidx}}{\slam{\kstr}{\svar{tmidx}}{\slam{\kargs}{\svar{args}}{\\
\quad \keyw{let}~\svar{aok} :: \kunit = \svar{arity0}~\svar{args}~\keyw{in}~\\
\quad \keyw{let}~\svar{rok} :: \kunit = \svar{rmatch}~\svar{tyidx}~\svar{tmidx}~\keyw{in}~\\
\quad \svar{str\_of\_Str}~\svar{tmidx}
}}}
\end{array}\]

Because regular strings are implemented as strings, this intro opcon definition is straightforward. It begins by making sure that no arguments were passed in (we will return to arguments and the fifth and sixth premises of (ana-intro) with the next example), using the helper function $\svar{arity0} :: \kargs \rightarrow \kunit$ defined such that any non-empty input will raise an error, via the static term $\sraise{\kunit}$. In practice, the tycon provider would specify an error message here.%The kinding rules and operational semantics of this exception-like form are standard:

Next, it checks the string provided as the term index against the regular expression given as the type index using $\svar{rmatch} :: \krx \rightarrow \keyw{Str} \rightarrow \kunit$, which we assume is defined in the usual way and again raises an error on failure. Finally, a {translation} corresponding to the term index is generated via  helper function $\svar{str\_of\_Str} :: \keyw{Str} \rightarrow \kitm$.% and $\svar{nat\_of\_Nat} :: \keyw{Nat} \rightarrow \kitm$ to generate  translational internal terms corresponding to the provided static terms and $\svar{len} :: \keyw{Str} \rightarrow \keyw{Nat}$ to statically compute the length of the string.

The only introductory form for kind $\kitm$ is $\sqitm{\qitm}$, where $\qitm$ is a \emph{translational internal term}. This form is analagous to the introductory form for kind $\kity$ described previously, $\sqity{\qity}$, where $\qity$ is a translational internal type. Each form in the syntax of $\iota$ has a corresponding form in the syntax for $\qitm$ and both the kinding rules and operational semantics simply recurse through these in the same manner as shown in Sec. \ref{sec:type-translations}. Also like translational internal types, there is an unquote form, $\quq{\st}$, that permits translational internal terms to be constructed compositionally. %The unquote form is eliminated in the course of evaluation. The supplement gives the complete rules.

The final two forms of translational internal term are $\anatrans{n}{\st}$ and $\syntrans{n}$. These represent the translation of argument $n$, the first if it arises via analysis against type $\st$ and the second if it arises by synthesis. They are not intended to be written directly by tycon providers, arising only internally in the semantics of argument interface lists. Before giving the rules, let us motivate the mechanism with the intro opcon definition for $\tcvar{lprod}$, $\omega_\text{lprod} := \tcstructn{\stx{lprod/intro}}; \omega_\text{lprod/targops}$ where $\stx{lprod/intro} :=$
\[\small
\begin{array}{l}
\lambda\svar{tyidx}{:}\klist{\klbl\times\kty}.\lambda\svar{tmidx}{:}\klist{\klbl}.\lambda\svar{args}{:}\kargs.\\
\quad \keyw{let}~\svar{inhabited}:\kunit=\svar{uniqmap}~\svar{tyidx}\\
\quad \skap{\kitm}{\skap{\karg}{\skap{\klbl}{\skap{\klbl\times\kty}{\svar{listrec3}}}}}~\svar{tyidx}~\svar{tmidx}~\svar{args}~\sqitm{\itriv}\\
\quad\quad \lambda \svar{rowtyidx}{:}\klbl\times\kty.\lambda \svar{rowtmidx}{:}\klbl.\lambda\svar{rowarg}{:}\karg.\lambda \svar{r}{:}\kitm.\\
\quad\quad\quad \keyw{letpair}~(\svar{rowlbl}, \svar{rowty})=\svar{rowtyidx}\\
\quad\quad\quad \keyw{let}~\svar{lok}::\kunit=\svar{lbleq}~\svar{rowlbl} ~\svar{rowtmidx}\\
\quad\quad\quad \keyw{let}~\svar{rowtr} :: \kitm = \svar{ana}~\svar{rowarg}~\svar{rowty}\\
\quad\quad\quad \sqitm{(\quq{\svar{rowtr}}, \quq{\svar{r}})
}
\end{array}
\]

The first line checks that the type provided is inhabited, in this case by checking that there are no duplicate labels via the helper function $\svar{uniqmap} :: \klist{\klbl \times \kty}\rightarrow \kunit$, raising an error if there are. An alternative strategy may have been to use an abstract kind that ensured that such type indices could not have been constructed, but to be compatible with our equality kind restriction, this would require support for abstract equality kinds, analagous to abstract equality types in SML. We chose not to formalize these for simplicity, and to demonstrate this more general technique: types with indices that are to be considered ``malformed'' can simply be left uninhabited. An analagous technique could be used to implement record types by requiring that the index be sorted.%, leaving indices where the labels did not appear sorted uninhabited. In both cases, we could provide a static helper function of kind $\klist{\klbl\times\kty}\rightarrow \kty$ that checked well-formedness immediately before constructing the requested type.

The rest of this opcon definition folds over the three lists provided as input: the list mapping labels to types provided as the type index, the list of labels provided as the term index, and the list of argument interfaces. We assume a straightforward helper function, $\svar{listfold3}$, that raises an error if the three lists are not of the same length. The base case is the translational empty product. The recursive case first checks that the label provided in the term index matches the label provided in the type index, using a helper function $\svar{lbleq} :: \klbl \rightarrow \klbl \rightarrow \kunit$. Then, we request type analysis of the corresponding argument, $\svar{rowarg}$,  against the type in the type index, $\svar{rowty}$, by writing $\svar{ana}~\svar{rowarg}~\svar{rowty}$, which  produces a translational internal term  referring to the translation of that argument if it succeeds, discussed next. The final line constructs a nested tuple based on this translation and the recursive result. Taken together, the translational internal term that will be derived for our example involving $e_\text{ex}$ above is $\small\qitm_\text{ex} := (\anatrans{0}{\stx{title}}, (\anatrans{1}{\stx{conf}}, ()))$.

\paragraph{Argument Interface Lists} We define $\karg$, the kind of \emph{argument interfaces}, as a simple product of functions, $\karg := (\kty \rightarrow \kitm) \times (\kunit \rightarrow \kty\times\kitm)$, one for analysis and the other for synthesis, both having the expected type. The helper functions $\svar{ana}$ and $\svar{syn}$ simply project the corresponding function out, $\svar{ana} := \slam{\karg}{\svar{arg}}{\keyw{fst}(\svar{arg})}$ and $\svar{syn}  :=  \slam{\karg}{\svar{arg}}{\keyw{snd}(\svar{arg})}$.

The argument interface list, called $\stx{args}$ in the rules, is simply a static list where the $i$th entry is a pair of functions of the form $(\slam{\kty}{\svar{ty}}{\sana{i}{\svar{ty}}}, \slam{\kunit}{\_}{\ssyn{i}})$. It is generated by the judgement $\keyw{args}(n)=\stx{args}$, where $n$ is given as the length of the argument list, written $|\es|=n$. See supplement.

%The $\kargs$ given where the $n$th entry is simply a pair of functions, the first of which allows for analysis of the $n$th argument of the operation against a type, and the second of which requests type synthesis for the $n$th argument. The helper functions simply project out the appropriate functions and invoke them.

 Recall that the kinding judgement is indexed by $n$, which is an upper bound on the argument index of such terms. This is enforced in the kinding rules:
\begin{mathpar}\small
\inferrule[k-ana]{
    n' < n\\
    \sofkX{\st}{\kty}
}{
    \sofkX{\sana{n'}{\st}}{\kitm}
}

\inferrule[k-syn]{
    n' < n
}{
    \sofkX{\ssyn{n'}}{\kty\times\kitm}
}
\end{mathpar}

The following lemma thus characterizes the argument interface list judgement:
\begin{lemma}
\small If $\keyw{args}(n)= \stx{args}$ then $\sofkn{\emptyset}{\emptyset}{\Phi}{n}{\stx{args}}{\kargs}$.
\end{lemma}

The rule (ocstruct-intro) ruled out writing either of these forms explicitly in an opcon definition by checking against the bound $n=0$. This is to prevent out-of-bounds errors -- tycon providers do not write these forms directly, but only access them via the argument interface list constructed by the semantics. This is possible because:
\begin{lemma}
\small If $\sofkn{\kDelta}{\kGamma}{\Phi}{n'}{\st}{\kappa}$ and $n > n'
$ then $\sofkn{\kDelta}{\kGamma}{\Phi}{n}{\st}{\kappa}$.
\end{lemma}

The dynamics for these forms couples the dynamics of the static language to the statics of the external language. For $\sana{n}{\st}$, after stepping the type $\st$ to a value and propagating errors, the argument environment, which stores the arguments themselves and the typing and tycon contexts, $\es; \Upsilon; \Phi$, is consulted to retrieve the $n$th argument and analyze it against $\st$. If this succeeds, the translational internal term $\sqitm{\anatrans{n}{\st}}$ is generated to refer to it.
\begin{mathpar}\small
\inferrule[s-ana-success]{
    \svalNA{\st}\\
    \keyw{nth}[n](\es) = e\\
    \eana{\Upsilon}{\Phi}{e}{\st}{\iota}
}{
    \sstep{\sana{n}{\st}}{\es;\Upsilon;\Phi}{\sqitm{\anatrans{n}{\st}}}
}
\end{mathpar}
If it fails, an error is raised:
\begin{mathpar}\small
\inferrule[s-ana-fail]{
    \svalNA{\st}\\
    \keyw{nth}[n](\es) = e\\
    [\Upsilon \vdash_\Phi e \nLeftarrow \st]
}{
    \serr{\sana{n}{\st}}{\es;\Upsilon;\Phi}
}
\end{mathpar}
We write $\small[\Upsilon \vdash_\Phi e \nLeftarrow \st]$ to indicate that $e$ fails to analyze against $\st$. Because we do not directly define this judgement  inductively, we also allow that this premise be left out, leaving a non-deterministic semantics, which does not affect our metatheory.

The dynamics for $\ssyn{n}$ are analagous, evaluating to a pair $(\st, \small\sqitm{\syntrans{n}})$ where $\st$ is the synthesized type.

The kinding rules also prevent the translational internal term forms generated by these static operations from being well-kinded when $n = 0$ so they can only arise in this manner (see supplement). Like the translational internal type form $\srep{\st}$, these  are retained in values of kind $\kitm$:
\begin{mathpar}\small
\inferrule[s-itm-anatrans-v]{
    \svalNA{\st}\\
    \keyw{nth}[n](\es)=e
}{
    \sval{\sqitm{\anatrans{n}{\st}}}{\es;\Upsilon;\Phi}
}

\inferrule[s-itm-syntrans-v]{
    \keyw{nth}[n](\es)=e
}{
    \sval{\sqitm{\syntrans{n}}}{\es;\Upsilon;\Phi}
}
\end{mathpar}

\paragraph{Abstracted Term Translations} The reason that the translations themselves were not inserted yet (even though they were derived in rule (s-ana-success) above) is again because we want to be able to  hold these abstract by replacing them with variables. The judgement $\edeabs{\tc}{\argEnv}{\qitm}{\memD}{\memG}{\itm}{\memD'}{\memG'}$, appearing as the eighth premise of (ana-intro), relates a translational internal term $\qitm$ to an internal term $\itm$ called the corresponding \emph{abstracted term translation}, where all references to the translation of an argument (having any type) are replaced with a fresh variable and all references to the translation of a type constructed by a user-defined tycon other than the ``delegated tycon'' $\tc$ are replaced with an abstract type variable as described in Sec. \ref{sec:selective-type-translation-abstraction}. The type translation store, $\memD$, discussed previously, and term translation store, $\memG$, track these mappings. Term translation stores have the following syntax: $\memG ::= \emptyset ~|~ \memG, n : \st \leadsto \iota/x : \tau$. Each entry can be read ``argument $n$ having type $\st$ and translation $\iota$ appears as variable $x$ with type $\tau$''. For example, \[\small\edeabs{\tcvar{lprod}}{(title, \texttt{"EXMPL 2015"}),\Upsilon,\Phi_\text{rstr}\Phi_\text{lprod}}{\qitm_\text{ex}}{\emptyset}{\emptyset}{\iota_\text{ex/abs}}{\memD_\text{ex/abs}}{\memG_\text{ex/abs}}\] where $\itm_\text{ex/abs} := (x_0, (x_1, ()))$, the term translation store is $\memG_\text{ex/abs} := 0 : \stx{title} \leadsto title/x_0 : \alpha_0, 1 : \stx{conf} \leadsto \texttt{"EXMPL 2015"}_\text{IL}/x_1 : \alpha_1$ and the type translation store $\memD_\text{ex/abs}$ is alpha-equivalent to $\memD_\text{paper/abs}$ from Sec. \ref{sec:selective-type-translation-abstraction}.

The rules for the abstracted term translation judgement follow recursively for shared forms just like those for the abstracted type translation judgement (see supplement). The rules for references to argument translations derived via type analysis operates as follows ($\syntrans{n}$ is analagous):
\begin{mathpar}\small
\inferrule[abs-anatrans-new]{
    n \notin \text{dom}(\memG)\\
    \keyw{nth}[n](\es) = e\\
    \eanaX{e}{\st}{\iota}\\
    \tdeabs{\Phi}{\tc}{\srep{\st}}{\memD}{\ity}{\memD'}\\
    (x~\text{fresh})
}{
    \edeabs{\tc}{\es;\Upsilon;\Phi}{\anatrans{n}{\st}}{\memD}{\memG}{x}{\memD'}{\memG, n : \st \leadsto \iota/x : \tau}
}
\end{mathpar}

Each $\memG$ induces an internal term substitution,  $\gamma ::= \emptyset ~|~ \gamma, \iota/x$, and corresponding internal typing context $\Gamma$ by the judgement $\memG \leadsto \gamma : \Gamma$, used as the ninth premise. In this case, $\small
\gamma_\text{ex/abs} := title/x_0, \texttt{"EXMPL 2015"}/x_1$ and $\small\Gamma_\text{ex/abs} := x_0 : \alpha_0, x_1 : \alpha_1$.

The tenth premise of (ana-intro) determines an abstract type translation for the type provided for analysis, in this case $\small\tau_\text{ex/abs} := \alpha_0 \times (\alpha_1 \times \kunit)$ (alpha-equivalent to $\tau_\text{conf/abs}$ in Sec. \ref{sec:selective-type-translation-abstraction}), and the eleventh premise extracts a type substitution, $\delta_\text{ex/abs}$, and type formation context, $\Delta_\text{ex/abs}$, from $\memD_\text{ex/abs}$, again equivalent to $\delta_\text{conf/abs}$ and $\Delta_\text{conf/abs}$ from Sec. \ref{sec:selective-type-translation-abstraction}.

Finally, the twelfth premise checks the abstracted term translation against the abstracted type translation. Here, we are checking $\Delta_\text{ex/abs}~\Gamma_\text{ex/abs} \vdash \iota_\text{ex/abs} : \tau_\text{ex/abs}$, i.e.: \[\small(\alpha_0, \alpha_1)~(x_0 : \alpha_0, x_1 : \alpha_1) \vdash (x_0, (x_1, ())) : \alpha_0 \times (\alpha_1 \times \kunit)
\]

In other words, the translation of the labeled product $e_\text{ex}$ generated by $\tcvar{lprod}$ is checked with all references to term and type translations of regular strings replaced by variables and type variables, respectively. Nevertheless, because our intro opcon definition treated arguments parametrically, the check succeeds. We will described  an ill-behaved operator, which would lead to a violation of the regular string tycon  invariant were it not disallowed by this check,  in Sec. \ref{metatheory}.

Applying the substitutions $\gamma_\text{ex/abs}$ and $\delta_\text{ex/abs}$ in the conclusion of the rule, we arrive at $\iota_\text{ex}$.

\subsection{Generalized Targeted Operations} 
All non-introductory opcons associated with user-defined tycons go through another generalized form, in this case for \emph{targeted operations}, $\etarg{\opname{op}}{\sttmidx}{e_\text{targ}}{\es}$, where $\opname{op}$ ranges over opcon names, ${\sttmidx}$ is the term index, $e_\text{targ}$ is the \emph{target argument} and $\es$ are the remaining arguments. Some examples of desugarings to this form include explicit invocation, $e_\text{targ}.\opname{op}\langle\sttmidx\rangle(\es)$, variants where the term index or the arguments are omitted, and label projection syntax for use by labeled product-like types, $e_\text{targ}\opname{\#}\conclbl{lbl}$, which desugars to $\etarg{\opname{\#}}{\conclbl{lbl}}{e_\text{targ}}{\cdot}$, and concatenative syntax $e_\text{targ} \cdot e_\text{arg}$, which desugars to $\etarg{\opname{conc}}{\striv}{e_\text{targ}}{e_\text{arg}}$. We show others, including case analysis, in the supplement.

Whereas introductory operations were strictly analytic, targeted operations are strictly synthetic in @$\lambda$. The type and translation is determined by the tycon of the type synthesized by the target argument (i.e. this tycon is used as the delegated tycon). The rule (syn-targ) is otherwise similar to (ana-intro) in its structure. The first premise synthesizes a type, $\sty{\tc}{\sttyidx}$, for the target argument. The second premise extracts the tycon definition for $\tc$ from the tycon context. The third extracts the \emph{operator index kind} from its opcon signature, and the fourth checks the term index against this kind. For example, we may associate the following targeted opcon signatures with the tycons $\tcvar{rstr}$ and $\tcvar{lprod}$:\[\small
\begin{array}{lcl}
\chi_\text{rstr/targops} & := & \opsig{\opname{conc}}{\kunit}, \opsig{\opname{case}}{\klist{\keyw{StrPattern}}}, \opsig{\opname{coerce}}{\krx}, \\
& & \opsig{\opname{check}}{\krx}, \opsig{\opname{replace}}{\krx}\\
\chi_\text{lprod/targops} & := & \opsig{\opname{prj}}{\klbl}, \opsig{\opname{conc}}{\klist{\klbl}}, \opsig{\opname{drop}}{\klist{\klbl}}
\end{array}
\]

The opcons associated with $\tcvar{rstr}$ are taken directly from Fulton et al.'s specification of regular string types \cite{sanitation-psp14}, with the exception of $\opname{case}$, which generalizes case analysis as defined there to arbitrary string patterns, which we discuss in the supplement. The opcons associated with $\tcvar{lprod}$ are also straightforward: $\opname{prj}$ projects out a labeled value from a labeled product, $\opname{conc}$ creates a new labeled product based on the target with some fields updated or added from the argument, and $\opname{drop}$ creates a new labeled product from the target with some fields dropped. Because targeted operations use no new mechanisms, we will only show regular string concatenation here. The others are in the supplement.

The fifth premise of (syn-targ) extracts the targeted opcon definition of $\opname{op}$ from the opcon structure, $\omega$. Like the intro opcon definition, this is a static function that generates a translational internal term on the basis of the delegated tycon's type index, the term index and an argument interface list. Targeted opcon definitions also synthesize a type. The rule (ocstruct-targ) in Fig. \ref{ocstruct} ensures that it is well-kinded. 
For example, $\omega_\text{rstr/targops}$ defines:
\vspace{-3px}
\[\small
\begin{array}{l}
\keyw{syn}~\opname{conc}=\slam{\krx}{\svar{tyidx}}{
    \slam{\kunit}{\svar{tmidx}}{
        \slam{\kargs}{\svar{args}}{\\
            \quad \keyw{letpair}~(\svar{arg1}, \svar{arg2}) = \svar{arity2}~\svar{args}\\
            \quad \keyw{letpair}~(\_, \svar{tr1}) = \svar{syn}~\svar{arg1}\\
            \quad \keyw{letpair}~(\svar{ty2}, \svar{tr2}) = \svar{syn}~\svar{arg2}\\
            \quad \stycase{\tcvar{rstr}}{\svar{ty2}}{\svar{tyidx2}}{\\
                \quad\quad (\sty{\tcvar{rstr}}{\svar{rxconcat}~\svar{tyidx}~\svar{tyidx2}}, \sqitm{sconcat~\quq{\svar{tr1}}~\quq{\svar{tr2}}})\\
                \quad
            }{\sraise{\kty\times\kitm}}
        }
    }
}
\end{array}
\]\vspace{-3px}
The first line simply checks that exactly two total arguments, including the target term, were passed in using the helper function $\svar{arity2}$. We then request synthesis of both arguments. We can ignore the type synthesized by the first because by definition it is a regular string type with type index $\svar{tyidx}$. We case analyze the second against $\tcvar{rstr}$, extracting its index regular expression if so and failing if not. We finally synthesize the resulting regular string type, using the helper function $\svar{rxconcat} :: \krx \rightarrow \krx \rightarrow \krx$ which generates the concatenated regular expression, and the translation, using an internal helper function $sconcat : \keyw{str} \rightarrow \keyw{str} \rightarrow \keyw{str}$, the translational internal term for which we assume has been substituted in directly above.

The remaining premises are identical to the corresponding premises in (ana-intro), with the only exception being that we check the abstract term translation against the abstract type translation of the synthesized type. In this example, only regular string types are involved so the abstracted type translations will not differ from the type translations themselves, i.e. $\keyw{str}$ (though abstracted arguments are still used to ensure hygiene, see Sec. \ref{sec:hygiene}).

\subsection{Other} The final form of external term is $\eother{\st}{\es}$. Rule (ana-other) specifies that it can be analyzed against $\sotherty{m}{\tau}$ if $\iota$ has type $\tau$ in the translation of the current typing context. We will see how it is used technically in the next section.
\begin{mathpar}\small

\end{mathpar}

% \section{Metatheory}\label{metatheory}
% %This judgement is only defined for values of kind $\kty$. We write $\st~\texttt{type}_\Phi$ iff $\vdash \Phi$ and $\sofkn{\emptyset}{\emptyset}{\Phi}{0}{\st}{\kty}$ and $\sval{\st}$.

% \begin{definition}[Argument Environment Formation] $\argEnvOK{n}{\es;\Upsilon;\Phi}$ iff $|\es|=n$ and $\vdash \Phi$ and $\vdash_\Phi \Upsilon \leadsto \Gamma$.\end{definition}

% \subsection{Static Language}

% %\begin{lemma}
% %If $\kDelta \vdash \kGamma$ and $\vdash \Phi$ and $\sofkn{\kDelta}{\kGamma}{\Phi}{n}{\st}{\kappa}$ then $\kDelta \vdash \kappa$.\todo{don't think we need this}
% %\end{lemma}

% \begin{theorem}[Static Canonical Forms] If $\sofkn{\emptyset}{\emptyset}{\Phi}{n}{\st}{\kappa}$ and $\argEnvOK{n}{\argEnv}$ and $\sval{\st}{\argEnv}$ then:
% \begin{enumerate}
% \item TODO: arrow
% \item TODO: unit
% \item TODO: product
% \item TODO: sum
% \item TODO: inductive
% \item TODO: universal
% \item If $\kappa = \kty$ then $\istype{\st}{\Phi}$ and 
%     \begin{enumerate}
%     \item $\st = \sty{\rightharpoonup}{(\st_1, \st_2)}$ and $\istype{\st_1}{\Phi}$ and $\istype{\st_2}{\Phi}$; or
%     \item $\st = \sty{\tc}{\sttyidx}$ and $\tcdef{\tc}{\tcsig{\ktyidx}{\chi}}{\omega} \in \Phi$ and $\sofkn{\emptyset}{\emptyset}{\Phi}{0}{\st}{\ktyidx}$ and $\svalNA{\sttyidx}$; or
%     \item $\st = \sotherty{m}{\tau}$ and $\emptyset \vdash \tau$.
%     \end{enumerate}
% \item If $\kappa = \kity$ then $\st=\sqity{\qity}$ and $\sofkn{\emptyset}{\emptyset}{\Phi}{0}{\st}{\kity}$ and $\svalNA{\st}$ and $\qtuq{\st'} \notin \qity$ and 
%     \begin{enumerate}
%     \item The outer form of $\qity$ is shared with $\tau$ and if $\qity'$ is a sub-term of $\qity$ then $\sofkn{\emptyset}{\emptyset}{\Phi}{0}{\sqity{\qity'}}{\kity}$ and $\svalNA{\sqity{\qity'}}$; or
%     \item $\qity = \srep{\st'}$ and $\istype{\st'}{\Phi}$
%     \end{enumerate}
% \item If $\kappa = \kitm$ then $\st=\sqitm{\qitm}$ and no sub-terms of $\qitm$ have the form $\quq{\st'}$ and 
%     \begin{enumerate}
%     \item The outer form of $\qitm$ is shared with $\itm$ and if $\qitm'$ is a sub-term of $\qitm$ then $\sofkn{\emptyset}{\emptyset}{\Phi}{n}{\sqitm{\qitm'}}{\kitm}$ and $\sval{\sqitm{\qitm'}}{\argEnv}$ and if $\qity$ is a sub-term of $\qitm$ then $\sofkn{\emptyset}{\emptyset}{\Phi}{0}{\sqity{\qity}}{\kity}$ and $\svalNA{\sqity{\qity}}$; or
%     \item $\qitm=\anatrans{n'}{\st'}$ and $n' < n$ and $\istype{\st'}{\Phi}$; or
%     \item $\qitm=\syntrans{n'}$ and $n' < n$.
%     \end{enumerate}
% \end{enumerate}
% \end{theorem}

% \begin{theorem}[Static Progress]
% If $\sofkn{\emptyset}{\emptyset}{\Phi}{n}{\st}{\kappa}$ and $\vdash \Phi$ and $\vdash^n \argEnv$ then $\sval{\st}{\argEnv}$ or $\serr{\st}{\argEnv}$ or $\sstep{\st}{\argEnv}{\st'}$.
% \end{theorem}
% \begin{proof}We proceed by rule induction on the kinding judgement.
% (k-parr)

% (k-ty)

% (k-otherty)

% (k-tycase-parr)

% (k-tycase)

% (k-ity-lam)

% (k-ity-alpha)

% (k-ity-unquote)

% (k-ity-trans)

% (k-raise)

% (k-itm-var)

% (k-itm-lam)

% (k-itm-unquote)

% (k-itm-anatrans)

% (k-itm-syntrans)
% \end{proof}

% \begin{theorem}[Static Preservation]
% If $\sofkn{\emptyset}{\emptyset}{\Phi}{n}{\st}{\kappa}$ and $\vdash \Phi$ and $\vdash^n \argEnv$ and $\sstep{\st}{\argEnv}{\st'}$ then $\sofkn{\emptyset}{\emptyset}{\Phi}{n}{\st'}{\kappa}$.
% \end{theorem}
% \begin{proof}
% (s-ty-step) (2 cases)
% (s-tycase-step) (2 cases)
% (s-tycase-match) (2 cases)
% (s-tycase-fail-1)
% (s-tycase-fail-2)
% (s-ity-lam-step-1)
% (s-ity-lam-step-2)
% (s-ity-unquote-step)
% (s-ity-unquote-elim)
% (s-ity-trans-step)
% (s-itm-lam-step-1)
% (s-itm-lam-step-2)
% (s-itm-unquote-step)
% (s-itm-unquote-elim)
% (s-ana-step)
% (s-ana-success)
% (s-syn-success) (relies on type synthesis theorem)
% (s-itm-anatrans-step)
% \end{proof}

% % \begin{lemma}[Kinding Stability]
% % If $\vdash \Phi$ and $\vdash \Phi, \tcdef{\tc}{\psi}{\theta}$ and $\kDelta \vdash \kGamma$ and $\sofkn{\kDelta}{\kGamma}{\Phi}{n}{\st}{\kappa}$ then $\sofkn{\kDelta}{\kGamma}{\Phi, \tcdef{\tc}{\psi}{\theta}}{n}{\st}{\kappa}$.
% % \end{lemma}
% % \begin{proof}
% % all cases straightforward by induction
% % \end{proof}

% % TODO: static normalization stability inside A?

% \subsection{Types}
% \begin{lemma}[Type Substitution Application]\label{thm:type-substitution-application}
% If $\vdash \delta : \Delta$ and $\Delta \vdash \tau$ then $\emptyset \vdash [\delta]\tau$.
% \end{lemma}

% \begin{lemma}[Selective Type Abstraction]\label{thm:selective-type-abstraction}
% If $\vdash \Phi$ and $\sofkz{\emptyset}{\emptyset}{\Phi}{\sqity{\qity}}{\kity}$ and $\svalNA{\sqity{\qity}}$ and $\memD \leadsto \delta : \Delta$ and $\vdash \delta : \Delta$ and $\tdeabs{\tc}{\Phi}{\qity}{\memD}{\tau}{\memD'}$ then $\memD' \leadsto \delta' : \Delta'$ and $\delta \subseteq \delta'$ and $\Delta \subseteq \Delta'$ and $\vdash \delta' : \Delta'$ and $\Delta' \vdash \tau$.
% \end{lemma}
% \begin{proof}
% (shared forms) (trans(st))
% \end{proof}

% \begin{lemma}[Type Translation]\label{type-translation}
% If $\vdash \Phi$ and $\istype{\st}{\Phi}$ and $\vdash_\Phi \st \leadsto \tau$ then $\emptyset \vdash \tau$.
% \end{lemma}
% \begin{proof} By Lemma \ref{thm:type-substitution-application} and Lemma \ref{thm:selective-type-abstraction}.
% \end{proof}

% \begin{lemma}[Typing Context Translation]
% If $\vdash \Phi$ and $\vdash_\Phi \Upsilon \leadsto \Gamma$ then $\emptyset \vdash \Gamma$.\end{lemma}
% \begin{proof} We proceed by rule induction on typing context translation. Empty case trivial. Extended case follows by Lemma \ref{type-translation} and definition of internal typing context formation.\end{proof}

% \subsection{External Terms}
% \begin{theorem}[Type Synthesis]
% If $\vdash \Phi$ and $\vdash_\Phi \Upsilon \leadsto \Gamma$ and $\esynX{e}{\st}{\iota}$ then $\istype{\st}{\Phi}$ and $\vdash_\Phi \st \leadsto \tau$. \end{theorem}
% \begin{proof}
% (var) (ap) (syn-targ)
% \end{proof}

% \begin{lemma}[Selective Term Abstraction]
% If $\vdash \Phi$ and $\vdash^n \argEnv$ and $\sofkn{\emptyset}{\emptyset}{\Phi}{n}{\sqitm{\qitm}}{\kitm}$ and $\sval{\sqitm{\qitm}}{\argEnv}$ and $\memD \leadsto \delta : \Delta$ and $\vdash \delta : \Delta$ and $\emptyset \vdash \Gamma_\text{out}$ and $\memG \leadsto \gamma : \Gamma$ and $\Delta~\Gamma_\text{out} \vdash \gamma : \Gamma$ and $\edeabs{\tc}{\argEnv}{\qitm}{\memD}{\memG}{\itm}{\memD'}{\memG'}$ then $\memD' \leadsto \delta' : \Delta'$ and $\memG' \leadsto \gamma' : \Gamma'$ and $\delta \subseteq \delta'$ and $\Delta \subseteq \Delta'$ and $\gamma \subseteq \gamma'$ and $\Gamma \subseteq \Gamma'$ and $\vdash \delta' : \Delta'$ and $\Delta'~\Gamma_\text{out} \vdash \gamma' : \Gamma'$.
% \end{lemma}
% \begin{proof} By rule induction on selective term abstraction judgement.

% (shared forms) (unquote form not possibly by canonical forms) (anatrans) (syntrans)
% \end{proof}

% \begin{theorem}[Type-Preserving Translation]
% If $\vdash \Phi$ and $\vdash_\Phi \Upsilon \leadsto \Gamma$ and either
% \begin{enumerate}
% \item $\eanaX{e}{\st}{\iota}$ and $\istype{\st}{\Phi}$; or 
% \item $\esynX{e}{\st}{\iota}$
% \end{enumerate}
% then $\vdash_\Phi \st \leadsto \tau$ then $\emptyset~\Gamma \vdash \iota : \tau$.
% \end{theorem}
% \begin{proof} By rule induction on typing rules.

% (var) (lam) (ap) (fix) (subsume) (ascribe) (ana-intro) (syn-targ) (other)
% \end{proof}

% TODO: stability of typing
% TODO: Unicity of typing
% TODO: hygiene?

% TODO: Make definitions out of these.
% The standard judgement $\Gamma \vdash \gamma : \Gamma'$ states that every binding $x : \tau$ in $\Gamma'$ has a corresponding substitution $\iota/x$ in $\gamma$ such that $\Gamma \vdash \iota : x$.

% , and  the judgement $\vdash \delta : \Delta$ checks that every type variable in $\Delta$ has a well-formed substitution in $\delta$

% \subsection{Conservativity}
% \begin{theorem}[Conservativity]
% If $\vdash \Phi$ and $\istype{\sty{\tc}{\sttyidx}}{\Phi}$ and for all $e$, if $\eana{\emptyset}{\Phi}{e}{\sty{\tc}{\sttyidx}}{\iota}$ and $\iota \Downarrow \iota'$ then $P(\iota')$, then if $\vdash \Phi, \tcdef{\tc'}{\psi}{\theta}$ then for all $e$, if $\eana{\emptyset}{\Phi, \tcdef{\tc'}{\psi}{\theta}}{e}{\sty{\tc}{\sttyidx}}{\iota}$ and $\iota \Downarrow \iota'$ then $P(\iota')$.
% \end{theorem}

% \begin{lemma}[Other Substitution]
% If $\vdash \Phi, \tcdef{\tc}{\psi}{\theta}$ and $\istype{\st}{\Phi}$ and $\eana{\Upsilon}{\Phi, \tcdef{\tc}{\psi}{\theta}}{e}{\st}{\iota}$ then there exists $e'$ such that $\eana{\Upsilon}{\Phi}{e'}{\st}{\iota}$.
% \end{lemma}

\section{Metatheory}\label{metatheory}
We will now state and sketch proofs of the key metatheoretic properties of @$\lambda$. The full proofs are in the supplement.

\paragraph{Type Safety}
Type safety requires that every well-typed external term translate to a well-typed internal term. Combined with type safety for the IL \cite{pfpl}, we have that well-typed external terms cannot go wrong. To prove this theorem, we must actually prove a stronger theorem, type-preserving translation, the analog of the notion of type-preserving compilation in typed compilation \cite{tarditi+:til-OLD}\todo{right citation for TIL?}.% produced as the translation of the external type. Note that we need only to state our top-level theorems for the analysis judgement  because of the subsumption rule.

\begin{theorem}[Type Preserving Translation]
If $\vdash \Phi$ and $\vdash_\Phi \Upsilon \leadsto \Gamma$ and $\istype{\st}{\Phi}$ and $\vdash_\Phi \st \leadsto \tau$ and $\eanaX{e}{\st}{\iota}$ then $\emptyset~\Gamma \vdash \iota : \tau$.
\end{theorem}
\begin{proof-sketch}
We proceed by induction on the typing judgement. The interesting cases are (ana-intro) and (syn-trans). The result follows directly from the final premise of these rules, combined with a lemma that states that the substitutions $\gamma$ and $\delta$ generated by the abstracted type and term translation stores satisfy $\Delta$ and $\Gamma$, and thus that applying them results in a well-typed term.
\end{proof-sketch}

\paragraph{Hygienic Translation} 
Note in rules (ana-intro) and (syn-trans) that the domains of $\Upsilon$ and $\Gamma$ are disjoint. This serves to ensure \emph{hygienic translation} -- translations cannot refer to variables in the surrounding scope directly, so uniformly renaming a variable cannot change the meaning of a program. Variables in $\Upsilon$ can (and do -- $title$ in the previous example) occur in arguments to the operation, but the translations of the arguments only appear \emph{after} the substitution $\gamma$ has been applied. We assume that applying a substitution is capture-avoiding in the usual manner. %(i.e. consistent with a locally nameless implementation).

\paragraph{Unicity}
The rules are structured so that if a term is well-typed, both its type and translation are unique. This can be stated as a unicity theorem.
\begin{theorem}[Unicity]
If $\vdash \Phi$ and $\vdash_\Phi \Upsilon \leadsto \Gamma$ and $\istype{\st}{\Phi}$ and $\istype{\st'}{\Phi}$ and $\eanaX{e}{\st}{\iota}$ and $\eanaX{e}{\st'}{\iota'}$ then $\st = \st'$ and $\iota = \iota'$.
\end{theorem}

\paragraph{Stability}
Extending the tycon context does not change the meaning of any terms that were previously well-typed.
\begin{theorem}[Stability]
Letting $\Phi' := \Phi, \tcdef{\tc}{\psi}{\theta}$, if $\vdash \Phi'$ and $\vdash_\Phi \Upsilon \leadsto \Gamma$ and $\istype{\st}{\Phi}$ and $\eanaX{e}{\st}{\iota}$ and $\vdash_\Phi \st \leadsto \tau$ then, $\vdash_{\Phi'} \Upsilon \leadsto \Gamma$ and $\istype{\st}{\Phi'}$ and $\eana{\Upsilon}{\Phi'}{e}{\st}{\iota}$ and $\vdash_{\Phi'} \st \leadsto \tau$.
\end{theorem}

\paragraph{Conservativity} 
Extending the tycon context also conserves all \emph{tycon invariants} maintained in any smaller tycon context. An example of a tycon invariant is the following:

\begin{tyconinvariant}[Regular Strings]
For all $e$ and $\sttyidx=\concrx{r}$, if $\eana{\emptyset}{\Phi_\text{rstr}}{e}{\sty{\tcvar{rstr}}{\sttyidx}}{\iota}$ and $\iota \Downarrow \iota'$ then $\iota'=\texttt{"s"}$ and $\texttt{"s"}$ is in the regular language $\mathcal{L}(\texttt{r})$.
\end{tyconinvariant}
\begin{proof-sketch} The proof of this invariant is not unusually difficult because we have fixed the tycon context $\Phi_\text{rstr}$, so we can simply reason about the language like a type-directed compiler specification for a calculus with only two tycons, $\rightharpoonup$ and $\tcvar{rstr}$. Such a calculus and compiler specification was given by Fulton et al. \cite{sanitation-psp14} and extending this to our  setting requires showing only that the opcon definitions in $\tcvar{rstr}$ adequately capture these specifications using standard program correctness techniques for the SL.\end{proof-sketch}

When we extend the tycon context, however, it is not trivially obvious that this invariant continues to hold. For example, a new tycon might define a targeted operator that synthesizes a regular string type, e.g. $\stx{paper}$, and generates a translation that is not in the corresponding regular language, e.g. $\texttt{""}$. The reason this cannot occur is because we hold translations of regular string types abstract  when passing through an opcon defined by any other tycon, and so, by standard parametricity properties for the IL, and the empty string does not have type $\alpha$. Indeed, no string other than one that came through an opcon associated with $\tcvar{rstr}$ can arise, so inductively, we know the invariant is maintained. We can state this more generally as a conservativity theorem.

\begin{theorem}[Conservativity] If $\vdash \Phi$ and $\tc \in \text{dom}(\Phi)$ and a tycon invariant for $\tc$ holds under $\Phi$: \begin{itemize}
\item for all $e$ and $\sttyidx$, if $\eana{\emptyset}{\Phi}{e}{\sty{\tc}{\sttyidx}}{\iota}$ and $\vdash_\Phi {\sty{\tc}{\sttyidx}} \leadsto \tau$ and $\iota \Downarrow \iota'$ then $P(\sttyidx, \iota')$ 
\end{itemize} then for all $\Phi' = \Phi, \tcdef{\tc'}{\psi'}{\theta'}$ such that $\vdash \Phi'$, the same tycon invariant holds under $\Phi'$: \begin{itemize}
\item for all $e$ and $\sttyidx$, if $\eana{\emptyset}{\Phi'}{e}{\sty{\tc}{\sttyidx}}{\iota}$ and $\vdash_{\Phi'}{\sty{\tc}{\sttyidx}}\leadsto{\tau}$ and $\iota \Downarrow \iota'$ then $P(\sttyidx, \iota)$.
\end{itemize}
\end{theorem}
\begin{proof-sketch}
For (ana-intro), we simply construct an equivalent introductory term that is well-typed in $\Phi$, where arguments that were of a type constructed by $\tc'$ are replaced by $\eother{\iota}$, making use of the fact that type $\sotherty{m}{\tau}$ is indistinguishable from a type outside $\Phi$ to the intro opcon definition for $\tc$ (see supplement for the full definition of this construction). For (syn-targ), we use the parametricity property described above if the synthesized type is constructed by $\tc$ to get our result by induction.
\end{proof-sketch}
%Had the translational term generated by the intro opcon definition been, e.g., $(\anatrans{0}{\stx{title}}, (\texttt{"TEST"}, ())$, then the corresponding abstract term would be $(x_0, (\texttt{"TEST"}, ())$ and the check would fail because $(\alpha_0, \alpha_1)~(x_0 : \alpha_0) \nvdash \texttt{"TEST"} : \alpha_1$. Had we not gone through the machinations of holding types and terms abstract, however, the check would succeed, because $\keyw{str}$ would no longer have been held abstract as $\alpha_1$. This form of ``type translation independence'' is precisely analagous to the representation independence properties that underly abstraction theorems for module systems based on  abstract types and will similarly serve to ensure that the type invariants maintained by $\tcvar{rstr}$ are conserved when new tycons are defined. In this case the invariant is that only strings that are in the regular language specified in the type index can be generated from an external term of regular string type. Note that $\texttt{"TEST"}$ is not in the regular language specified by $\concrx{[A-Z]+ \digit\digit\digit\digit}$, so it is critical that it not be allowed, despite being consistent with the type translation of $\stx{conf}$, i.e. it is a string. Such an invariant could not be maintained if the type system were treated as merely a ``bag of rules''.\todo{put conservativity here?} %We consider this more rigorously in Sec. \ref{metatheory}.

\section{Related Work and Discussion}\label{prior-work}

% Note also that type constructors are not themselves static values. To apply them in a curried manner or pass them around in the static language, we can define a static function, e.g. $\slam{\kty}{\svar{t1}}{\slam{\kty}{\svar{t2}}{\sparr{(\svar{t1}, \svar{t2})}}}$ has kind $\karrow{\kty}{\karrow{\kty}{\kty}}$.

%To situate our work, and because we build on some recent results, we begin by summarizing prior approaches that can be considered when a simple isomorphic embedding of a fragment is not evident.

\paragraph{Term Rewriting}
Language-integrated static term rewriting  (``macro'') systems, like Template Haskell \cite{SheardPeytonJones:Haskell-02} and Scala's static macros \cite{ScalaMacros2013}, can be used to decrease complexity when an isomorphic embedding into the underlying type system is possible, but cannot extend the type system directly.% If each metaprogram is invoked like a function, inner macros are expanded before outer macros, and the mechanism enforces abstraction barriers by ensuring that the rewriting logic is hygienic, cannot modify surrounding code, and does not depend on shared state, then compositional reasoning is possible: the meaning of a term depends only on its subterms, not on the specific position it appears in a program.% Complex macros can, however, still be difficult to reason about, so many static macro systems impose further constraints (e.g. Scala recommends using only its ``black box'' macros, which  enforce a function-like typing discipline). These do not permit extensions to the underlying type system.
%Rewriting systems that permit global pattern-based dispatch, however, do not admit strong modular reasoning principles (the same term might match multiple patterns, defined separately, creating ambiguities).

% \paragraph{Desugaring}\label{desugaring}
% Sometimes, dialects address issues of complexity by introducing new syntax. Indeed, most dialects do build in syntax for a few privileged abstractions (e.g. list literals are nearly ubiquitous, monad comprehensions support a key feature of Haskell). %For example, even though lists can be defined using datatypes, most languages build in special ``literal forms'' that make introducing and pattern matching over lists less tedious. Different dialects may similarly choose to  provide concise forms for working with other types of data (e.g. option types, regular expressions or SQL queries). Types not in the language's standard library, however, must typically use a more uniform syntax. 
% %Evidence suggests that this leads programmers to choose less semantically meaningful representations (e.g. representing SQL queries as strings), which can cause a number of issues \cite{TSLs}. 
% To make the situation less asymmetric, systems that introduce new  ``desugarings'' atop a base language have been developed. For example, %Camlp4 is a language-external system used by the OCaml community. 
% dialects of languages built using Sugar* \cite{erdweg2013framework} allow syntax extensions to be packaged separately and combined using language-integrated declarations. If the desugaring logic obeys the same constraints described above for macros, semantic reasoning can be performed compositionally (and recent work has shown how reasoning about type correctness can be automated \cite{conf/icfp/LorenzenE13}).

% These systems do not guarantee that the composition of unambiguous grammars will remain unambiguous (and realistic examples of ambiguous combinations are not uncommon, e.g. XML and HTML). %Every combination of desugarings is thus a  dialect with respect to syntactic reasoning. 
% Recent work has made progress in addressing this problem by restricting how syntax extensions can interface with the host syntax. Schwerdfeger and Van Wyk require a globally unique start token and describe checkable conditions pertaining to the follow sets of host language non-terminals to guarantee that extensions can be composed unambiguously \cite{conf/pldi/SchwerdfegerW09}. This suffices for modularly adding new keyword-prefixed forms, but literal forms are awkward to define in this way. Omar et al. describe a language-integrated technique specifically for this purpose, using a technique with parallels to the one we are building on \cite{TSLs}. Literal parsing logic is directly associated with user-defined types, forming \emph{type-specific languages} (TSLs), and local type inference controls invocation of TSL parsing logic, guaranteeing that composition is unambiguous. The mechanism guarantees hygiene and inner terms cannot be inspected, so  fully modular reasoning is possible.

%We assumed these desugaring techniques were available to solve issues strictly related to complexity and convenience.

\paragraph{Optimization and Extensible Compilers}
When a simple embedding that preserves the static semantics exists, but a different  embedding would better preserve a desired cost semantics, term rewriting techniques can also be used to perform ``optimizations'', thus achieving an isomorphic embedding. Care must be taken, however, to ensure that the optimized value is not manipulated directly if the rewriting does not preserve the static semantics (i.e. the rewritten term has a different, less constrained type). Type abstraction has been used directly to achieve this property in work on \emph{lightweight modular staging} \cite{Rompf:2012:LMS}. This does not allow new type systems to be defined (in LMS, Scala's type system). %As long as transformations are meaning-preserving (which can also be shown mechanically), they can be applied in any order. %The metaprograms can modularly be shown not to violate the type system of the host language, Modular reasoning is possible as long as dispatch is explicit (e.g. macros) or based on an unambiguous language mechanism (e.g. trait composition in LMS) and made simpler by the fact that transformations must be type-preserving.

\paragraph{Type Refinements}
%When T, the solution is less clear. 
When new static distinctions need to be made for values of an existing type, but new primitive operations are not needed, one solution is to develop a system of \emph{type refinements} based on the fragment of interest \cite{Freeman91}. For example, one might refine the type of integers to distinguish negative integers. Most proposals for ``pluggable type systems'' describe such type refinement systems \cite{Brac04a}, which require only additional annotations supplied as comments or metadata (e.g. JavaCOP \cite{Andreae:2006:FIP:1167473.1167479}). Refinement systems do not support holding the refined type abstract as we did in our regular string example ($\tcvar{rstr}$ could have picked a non-string representation for, e.g., fixed size regular languages)%Because the dynamic semantics are fixed,  composition is safe. Although our regular string types could perhaps be approximated with refinements, this does not permit the advanced operations, optimizations and representation independence results we enable.

\paragraph{Language Frameworks}
The most general situation, of which type refinements are a special case, is when exposing a fragment requires defining new types as well as new operators, where the static and dynamic semantics governing these new operators make non-trivial use of  statically valued information. 
We saw labeled product types added, which require a new type constructor and new operator constructors.% The regular string type's group projection operator had a specialized cost semantics, so a simple refinement would not be able to approximate it. Scala's type system does not track the necessary invariants, so LMS-like optimizations would also not be appropriate.%For example, projection operators,  written concretely in some dialects of ML as \verb|#label|, can be seen as applications of \verb|#|, the projection operator constructor, to \verb|label|, a static index. The type of an operation like \verb|#label e| (sometimes written postfix as \verb|e#label|) is a function of the index of the type of \verb|e|. The dynamics also needs (some trace of) this information to determine how the operation is evaluated.

%Many of the other examples mentioned above as motivations for new dialects have a similar flavor. We will detail another example in the next section of a type constructor that tracks strings known statically to be in a regular language, using it to dispense with unnecessary run-time checks in cases where they are not strictly necessary (affecting the cost component of the dynamic semantics).

%An important observation is that while it is possible to \emph{implement} records by translation to tuples, or nested binary pairs, or lists, or many other types arising from a language like $\mathcal{L}\{{\rightharpoonup}\,{\forall}\,{\mu}\,{\keyw{1}}\,{\times}\,{+}\}$, it is not possible to isomorphically embed them as such (and thus rely on purely syntactic sugar) because there is no unique inverse mapping from, e.g., unlabeled product types to the corresponding record type (many combinations of labeled products may be implemented using the same unlabeled product). This is a common pattern even in more complex situations: a simple internal language (IL) generally suffices as a target for implementing the dynamics of an external language (EL) with a richer static semantics, but an embedding is not possible. This observation underlies the design of most compilers.

 % like SML, Specifying languages in a manner analagous to how they are usually implemented can simplify reasoning about the core of the language.

% and typed translation semantics (the internal terms have a different type system).% Translation semantics

%These types do not admit an isomorphic embedding using only unlabeled tuples because the field projection family of operators, indexed by labels (field projection is written concretely as \verb|#label(e)|, \verb|e#label| or \verb|e.label| in different languages), must be capable of operating on all values of such types. % with the labels removed would not admit a unique inverse. 
%(each equivalence class of record types could be implemented as an abstract type using tuples at considerable inconvenience but there would still be no way to define a projection operator that operated uniformly over all records having a particular field).

% (i.e. fragments defining families of operations that cannot simply be written as a finite collection of functions, examples of which we will discuss). If there are a finite number of operations, and they can be typechecked without inspecting the structure of the types of the arguments, then an abstract type can be used. new primitive operations (including new variants of existing operations) that require static knowledge of these invariants, however, this too is infeasible. take advantage of the invariants being maintained, and these operations must be usable throughout a program, . %At best, annotations might be written as ``comments'' or metadata for use by these tools.

A variety of \emph{language frameworks} have been developed to make it easier to define new languages, in some cases together with a compiler and other tooling for these languages; see \cite{erdweg2013state}\todo{unlimited references so add a bunch here}. %Some \emph{extensible compilers} are also actually language frameworks, because they permit the introduction of new constructs, not just new meaning-preserving optimizations. The maintainers of many compilers, e.g. Haskell's GHC, treat their compiler as a ``laboratory'' for new fragments, toggled by flags or special comments. \emph{Logical frameworks} focus more specifically on helping language designers mechanize the metatheory of language and logic specifications. T
These tools can decrease the effort needed to define a new language, but they either do not support forming languages from separately defined fragments or provide few modular reasoning principles that make it possible to reason separately about these fragments. %Every combination of fragments is a new dialect, and must be reasoned about monolithically. %Efforts to reason modularly areeed, several problems can come up when fragments are combined, so . 
% Let us briefly review difficulties that arise.

% %Concrete syntax known to be separately unambiguous might not be unambiguous when combined, as was already discussed above. 
% If the abstract syntax needs to be extended to support a new fragment, problems also arise. In monolithic settings, terms can be implemented using finite recursive sums (i.e. term constructors are often implemented as ML-style datatype constructors), but this does not permit extension, so open sum types or products of functions (i.e. objects \cite{conf/oopsla/Aldrich13}) must instead be used. This can present issues when one wishes to modularly define new functionality that should exhaustively cover all terms in the language (e.g. pretty-printers for expressions). Reynolds first identified this problem \cite{Reynolds75} and Wadler named it the \emph{expression problem}. 
% %A number of language frameworks (e.g. JastAdd \cite{Ekman:2007:JEJ:1297027.1297029} and Silver \cite{VanWyk:2010:SEA}) use extensible \emph{attribute grammars}. These can be seen as a form of open sum, where attributes correspond to functions performing traversals (more specifically, \emph{catamorphisms} \cite{catamorphisms}\todo{citation / remove?}). %When a new term constructor is added, the logic determining how some existing attributes (e.g. typechecking and translation) should handle the new case is also provided. \emph{Forwarding} can be used to attempt to delegate responsibility over attributes other than those explicitly defined to another term constructor, eventually leading to one in the fixed internal language \cite{VanWyk:2010:SEA}. This can address some aspects of the expression problem, but creates a bigger problem: the internal implementation details  of a fragment are necessarily exposed, violating an abstraction barrier that, as we will discuss, is critical.
% In this work, we sidestep the problems of syntax, instead leaving it fixed and relying on a bidirectional type system delegating to a relevant tycon. %For example, projection operators do not require adding a corresponding term constructor to the abstract syntax. Instead, projection is categorized as a \emph{targeted operation}, so  the concrete term \verb|#label e| desugars to an abstract term like $\keyw{targ}[(\desugar{\conclbl{\#}}, \desugar{\conclbl{\texttt{label}}})](e)$ (where $\desugar{\cdot}$ denotes an encoding of constant labels, here a label corresponding to the operator constructor and the field label itself, into the static language). The type constructor of the type synthesized by the target argument, here $e$, determines the semantics of the term. If a new record-like fragment is added (e.g. they give the example of one with prototypic inheritance), it can reuse the same concrete and abstract syntax directly, precluding conflicts and avoiding the need to define the behavior of tools like pretty printers for every new fragment. As in their work on type-specific languages, described above, the semantics of literal forms are controlled by the type constructor of the type the literal is being analyzed against. 
% %We will discuss this in the next section.

% Once syntactic issues have been addressed, however, there are a host of semantic guarantees  that must be established before a language can be relied upon, as we saw. % The most basic guarantee is \emph{type safety}: that the dynamics are well-defined for all terms accepted as well-typed by the statics, and preserve the statics during evaluation. Each fragment can then build on type safety to establish additional \emph{type invariants} stating properties about all types that it constructs, which  clients can then use to reason about programs (e.g. value induction for eager sum types relies on finiteness of the canonical forms). In practice, typechecking is expected be \emph{deterministic} and \emph{decidable} (i.e. terminating) and, except in circumstances where non-determinism is explicitly exposed to programmers, the dynamic semantics are also expected to be deterministic. %Precise formulations of these properties depend on how the semantics are specified.%, so we will make this more precise below.
% Modern language frameworks guarantee few or none of these properties about the dialects they produce. More alarmingly, even when these properties have been established for two dialects  (either in the metatheory or mechanically using a logical framework), and syntactic conflicts are addressed, there is no guarantee that merging the dialects together will conserve these properties. % If our goal is to integrate fragment composition into the language, these tools are thus of limited utility. Clients would have to take on the burden of reasoning about these basic properties for every combination of ``libraries'' they chose to import.% Improving this state of affairs, so that it more resembles reasoning about separately defined modules (we assume an ML-style module system), is the general topic of this paper.

% %Though we have not mechanically proven these properties for our design, we have given strong evidence that these properties are maintained modularly. That is, we need only establish the semantics of the opcon structures in isolation. %We can then rely on, e.g., the Conservativity theorem above.

%We note that is increasingly being deployed in contemporary languages (e.g. Scala) due to the perception that it enables good error messages, and to support semantics where  non-local type inference is undecidable. Permitting extensibility together with non-local type inference is an open problem.
\paragraph{Module Systems}
Our work has some technical similarities to work on maintaining type abstraction using a form of effect system \cite{CraryHarperDreyer2002}, and on translating modules to System F \cite{conf/tldi/RossbergRD10}. Unlike modules, tycons are indexed by arbitrary static values, permit the representation of a type to be computed based on this index, and compute both the type and translation of each operation, rather than requiring that the operations be expressible as a finite collection of functions.% This permits us tocode targeting a separate IL, and permit operations whose types cannot simply be written as functions. %Both examples we used here would be quite difficult to fully embed using only modules.

\section{Discussion}\label{discussion}
%The most immediate direction for future work is to finishing mechanizing the metatheory for this  technique, and to embed it into an existing proof assistant (using a continuation passing style to simulate the stores in our normalization semantics). Despite this, we believe that the technique as described semi-formally above is compelling. Adding support for tycon-specific typing contexts, implicit coercions and direct opcon calls can all be taken up by the community. Delving further into the question of when can two tycons with the same signature be substituted for one another, using a technique based on admissible relations, is also an avenue we wish to explore \cite{pfpl}.

%We combine several interesting type theoretic techniques, applying them to novel ends: 1)  a bidirectional type system permits flexible reuse of a fixed syntax; 2) the SL serves both as an extension language and as the type-level language; we give it its own statics (i.e. a \emph{kind system}); 3) we use a typed intermediate language and leverage corresponding \emph{typed compilation} techniques, here lifted into the semantics of the EL; 4) we leverage internal type abstraction implicitly as an effect during normalization of the SL to enforce abstraction barriers between type constructors. 
%As a result, conservativity follows from the same elegant parametricity results that underly  abstraction theorems for module systems. 
%Like modules, reasoning about these \emph{modular type constructors} does not require  mechanized specifications or proofs: correctness issues in the type constructor logic necessarily causes typechecking to fail, so even extensions that are not proven correct can be distributed and ``tested'' in the wild without compromising the integrity of an entire program (at worst, only values of types constructed by the tycon being tested may exhibit undesirable properties).

%\acks
%The author is grateful to Jonathan Aldrich, Robert Bocchino and anonymous referees for their useful suggestions. This work was funded by the DOE Computational Science Graduate Fellowship under grant number DE-FG02-97ER25308.
%Acknowledgments, if needed.

\newpage
% We recommend abbrvnat bibliography style.

\bibliographystyle{abbrv}

% The bibliography should be embedded for final submission.

\bibliography{../research}
%\softraggedright
%P. Q. Smith, and X. Y. Jones. ...reference text...

\newpage
\appendix
\section{Appendix}
\begin{mathpar}
\small
\inferrule[s-ty-step]{
    \sstep{\st}{\argEnv}{\st'}
}{
    \sstep{\sty{c}{\st}}{\argEnv}{\sty{c}{\st'}}
}

\inferrule[s-ty-err]{
    \serr{\st}{\argEnv}
}{
    \serr{\sty{c}{\st}}{\argEnv}
}

\inferrule[s-ty-v]{
    \sval{\st}{\argEnv}
}{
    \sval{\sty{c}{\st}}{\argEnv}
}

\inferrule[s-otherty-v]{ }{
    \sval{\sotherty{m}{\tau}}{\argEnv}
}
\end{mathpar}

\begin{mathpar}
\small
\inferrule[s-tycase-step]{
    \sstep{\st}{\argEnv}{\st'}
}{
    \sstep{\stycase{c}{\st}{\sx}{\st_1}{\st_2}}{\argEnv}{\stycase{c}{\st'}{x}{\st_1}{\st_2}}
}

\inferrule[s-tycase-err]{
    \serr{\st}{\argEnv}
}{
    \serr{\stycase{c}{\st}{\sx}{\st_1}{\st_2}}{\argEnv}
}
\end{mathpar}
\begin{mathpar}\small
\inferrule[s-tycase-match]{
    \sval{\sty{c}{\st}}{\argEnv}
}{
    \sstep{\stycase{c}{\sty{c}{\st}}{\sx}{\st_1}{\st_2}}{\argEnv}{[\st/x]\st_1}
}

\inferrule[s-tycase-fail]{
    \st \neq \sty{c}{\st'}
}{
    \sstep{\stycase{c}{\st}{\sx}{\st_1}{\st_2}}{\argEnv}{\st_2}
}
\end{mathpar}
\begin{mathpar}
\small
\inferrule[keq-k]{\kvar \in \kDelta}{\keq{\kDelta}{\kvar}}

\inferrule[keq-ind]{\keq{\kDelta, \kvar}{\kappa}}{
  \keq{\kDelta}{\kmu{\kvar}{\kappa}}}

\inferrule[keq-unit]{ }{\keq{\kDelta}{\kunit}}
\end{mathpar}\begin{mathpar}
\small
\inferrule[keq-prod]{
  \keq{\kDelta}{\kappa_1}~~~~
  \keq{\kDelta}{\kappa_2}
}{
  \keq{\kDelta}{\kprod{\kappa_1}{\kappa_2}}
}

\inferrule[keq-sum]{
  \keq{\kDelta}{\kappa_1}~~~~
  \keq{\kDelta}{\kappa_2}
}{
  \keq{\kDelta}{\ksum{\kappa_1}{\kappa_2}}
}

\inferrule[keq-ty]{ }{\keq{\kDelta}{\kty}}
\end{mathpar}
 %We will discuss strategies for fragments that benefit from more sophisticated equivalences in Section \ref{examples}.\todo{do this}

\begin{mathpar}

\inferrule[k-ity-alpha]{ }{\sofkX{\sqity{\alpha}}{\kity}}
\end{mathpar}

\begin{mathpar}
\small
\inferrule[s-ity-lam-step-1]{
    \sstep{\sqity{\qity_1}}{\argEnv}{\sqity{\qity_1'}}%\sstep{\sqity{\qity_1}}{\argEnv}{\sstep{\sqity{\qity_1'}}}
}{
    \sstep{\sqity{\qity_1\times\qity_2}}{\argEnv}{\sqity{\qity_1' \times \qity_2}}
}

\inferrule[s-ity-lam-step-2]{
    \sval{\sqity{\qity_1}}{\argEnv}\\
    \sstep{\sqity{\qity_2}}{\argEnv}{\sqity{\qity_2'}}%\sstep{\sqity{\qity_1}}{\argEnv}{\sstep{\sqity{\qity_1'}}}
}{
    \sstep{\sqity{\qity_1\times\qity_2}}{\argEnv}{\sqity{\qity_1 \times \qity_2'}}
}

\inferrule[s-ity-lam-err-1]{
    \serr{\sqity{\qity_1}}{\argEnv}
}{
    \serr{\sqity{\qity_1\times\qity_2}}{\argEnv}
}

\inferrule[s-ity-lam-err-2]{
    \serr{\sqity{\qity_2}}{\argEnv}
}{
    \serr{\sqity{\qity_1\times\qity_2}}{\argEnv}
}

\inferrule[s-ity-lam-v]{
    \sval{\sqity{\qity_1}}{\argEnv}\\
    \sval{\sqity{\qity_2}}{\argEnv}
}{
    \sval{\sqity{\qity_1\times\qity_2}}{\argEnv}
}

\inferrule[s-ity-alpha-v]{ }{
    \sval{\sqity{\alpha}}{\argEnv}
}
\end{mathpar}
\begin{mathpar}
\small
\inferrule[k-tycase-parr]{
    \sofkX{\st}{\kty}\\
    \sofk{\kDelta}{\kGamma, \sx :: \kty \times \kty}{\Phi}{\st_1}{\kappa}\\
    \sofkX{\st_2}{\kappa}
}{
    \sofkX{\stycase{\rightharpoonup}{\st}{\sx}{\st_1}{\st_2}}{\kappa}
}
\end{mathpar}
\begin{mathpar}
\small
\inferrule[s-ity-unquote-step]{
    \sstep{\st}{\argEnv}{\st'}
}{
    \sstep{\sqity{\qtuq{\st}}}{\argEnv}{\sqity{\qtuq{\st'}}}
}

\inferrule[s-ity-unquote-err]{
    \serr{\st}{\argEnv}
}{
    \serr{\sqity{\qtuq{\st}}}{\argEnv}
}
\end{mathpar}
\begin{mathpar}
\inferrule[s-ity-trans-step]{
    \sstep{\st}{\argEnv}{\st'}
}{
    \sstep{\sqity{\srep{\st}}}{\argEnv}{\sqity{\srep{\st'}}}
}

\inferrule[s-ity-trans-err]{
    \serr{\st}{\argEnv}
}{
    \serr{\sqity{\srep{\st}}}{\argEnv}
}
\end{mathpar}
This judgement is defined by the following straightforward rules:
\begin{mathpar}\small
\inferrule[tstore-emp]{ }{\emptyset \leadsto \emptyset : \emptyset}

\inferrule[tstore-ext]{\memD \leadsto \delta : \Delta}{(\memD, \st \leftrightarrow \tau/\alpha) \leadsto (\delta, \tau/\alpha) : (\Delta, \alpha)}
\end{mathpar}

\[\small
\begin{array}{lll}
\textbf{Description} ~~~~& \textbf{Concrete Form}~~~~ & \textbf{Desugared Form}\\
\text{sequences} & (e_1, \ldots, e_n) ~\text{or}~ \texttt{[}e_1, \ldots, e_n\texttt{]} & \eintro{()}{e_1; \ldots; e_n}\\
\text{labeled sequences} & \{\mathtt{lbl}_1=e_1, \ldots, \mathtt{lbl}_n=e_n\} & \eintro{\texttt{[}\mathtt{lbl}_1, \ldots, \mathtt{lbl}_n\texttt{]}}{e_1; \ldots; e_n}\\
\text{label application} & \mathtt{lbl}\langle e_1, \ldots, e_n \rangle & \eintro{\mathtt{lbl}}{e_1, \ldots, e_n}\\
\text{numerals} & n & \eintro{n}{\cdot}\\
\text{labeled numerals} & n\texttt{lbl} & \eintro{(n, \texttt{lbl})}{\cdot}\\
\text{strings} & \texttt{"s"} & \eintro{\texttt{"s"}}{\cdot}
\end{array}
\]

\begin{mathpar}\small
\inferrule[s-itm-var-v]{ }{
    \sval{\sqitm{x}}{\argEnv}
}

\inferrule[s-itm-lam-step-1]{
    \sstep{\sqity{\qity}}{\argEnv}{\sqity{\qity'}}
}{
    \sstep{\sqitm{\ilam{\qity}{x}{\qitm}}}{\argEnv}{\sqitm{\ilam{\qity'}{x}{\qitm}}}
}

\inferrule[s-itm-lam-step-2]{
    \sval{\sqity{\qity}}{\argEnv}\\
    \sstep{\sqitm{\qitm}}{\argEnv}{\sqitm{\qitm'}}
}{
    \sstep{\sqitm{\ilam{\qity}{x}{\qitm}}}{\argEnv}{\sqitm{\ilam{\qity}{x}{\qitm'}}}
}
\end{mathpar}
\begin{mathpar}\small
\inferrule[s-itm-lam-err-1]{
    \serr{\sqity{\qity}}{\argEnv}
}{
    \serr{\sqitm{\ilam{\qity}{x}{\qitm}}}{\argEnv}
}

\inferrule[s-itm-lam-err-2]{
    \serr{\sqitm{\qitm}}{\argEnv}
}{
    \serr{\sqitm{\ilam{\qity}{x}{\qitm}}}{\argEnv}
}

\inferrule[s-itm-lam-v]{
    \sval{\sqity{\qity}}{\argEnv}\\
    \sval{\sqitm{\qitm}}{\argEnv}
}{
    \sval{\sqitm{\ilam{\qity}{x}{\qitm}}}{\argEnv}
}
\end{mathpar}
\begin{mathpar}\small
\inferrule[k-itm-unquote]{
    \sofkX{\st}{\kitm}
}{
    \sofkX{\sqitm{\quq{\st}}}{\kitm}
}
\end{mathpar}

\begin{mathpar}\small
\inferrule[s-itm-unquote-step]{
    \sstep{\st}{\argEnv}{\st'}
}{
    \sstep{\sqitm{\quq{\st}}}{\argEnv}{\sqitm{\quq{\st'}}}
}

\inferrule[s-itm-unquote-err]{
    \serr{\st}{\argEnv}
}{
    \serr{\sqitm{\quq{\st}}}{\argEnv}
}

\inferrule[s-itm-unquote-elim]{
    \sval{\sqitm{\qitm}}{\argEnv}
}{
    \sstep{\sqitm{\quq{\sqitm{\qitm}}}}{\argEnv}{\sqitm{\qitm}}
}
\end{mathpar}
\begin{mathpar}
\inferrule[s-ana-step]{
    \sstep{\st}{\argEnv}{\st'}
}{
    \sstep{\sana{n}{\st}}{\argEnv}{\sana{n}{\st'}}
}

\inferrule[s-ana-err]{
    \serr{\st}{\argEnv}
}{
    \serr{\sana{n}{\st}}{\argEnv}
}

\inferrule[s-itm-anatrans-step]{
    \sstep{\st}{\argEnv}{\st'}
}{
    \sstep{\sqitm{\anatrans{n}{\st}}}{\argEnv}{\sqitm{\anatrans{n}{\st'}}}
}

\inferrule[s-itm-anatrans-err]{
    \serr{\st}{\argEnv}
}{
    \serr{\sqitm{\anatrans{n}{\st}}}{\argEnv}
}
\end{mathpar}
, as specified by the judgement $\memG \leadsto \gamma : \Gamma$ defined by the following rules:
\begin{mathpar}
\inferrule[ttrs-emp]{ }{\emptyset \leadsto \emptyset : \emptyset}

\inferrule[ttrs-ext]{
    \memG \leadsto \gamma : \Gamma
}{
    (\memG, n : \st \leadsto \iota/x : \tau) \leadsto (\gamma, \iota/x) : (\Gamma, x : \tau)
}
\end{mathpar}

\begin{mathpar}\small
\inferrule[k-itm-lam]{
    \sofkX{\sqity{\qity}}{\kity}\\
    \sofkX{\sqitm{\qitm}}{\kitm}
}{
    \sofkX{\sqitm{\ilam{\qity}{x}{\qitm}}}{\kitm}
}
\end{mathpar} 
\begin{mathpar}\small
\inferrule[k-raise]{\kDelta \vdash \kappa}{\sofkX{\sraise{\kappa}}{\kappa}}

\inferrule[s-raise]{ }{\serr{\sraise{\kappa}}{\argEnv}}
\end{mathpar}

\begin{mathpar}\small
\inferrule[s-syn-success]{
    \keyw{nth}[n](\es) = e\\
    \esynX{e}{\st}{\iota}
}{
    \sstep{\ssyn{n}}{\es;\Upsilon;\Phi}{(\st, \sqitm{\syntrans{n}})}
}

\inferrule[s-syn-fail]{
    \keyw{nth}[n](\es) = e\\
    [\Upsilon \vdash_\Phi e \nRightarrow]
}{
    \serr{\ssyn{n}}{\es;\Upsilon;\Phi}
}
\end{mathpar}

\begin{mathpar}\small
\inferrule[k-itm-syntrans]{
    n' < n
}{
    \sofkX{\sqitm{\syntrans{n'}}}{\kitm}
}
\end{mathpar}
, e.g. for lambdas:
\begin{mathpar}\small
%\inferrule[abs-var]{ }{\edeabs{\tc}{\argEnv}{x}{\memD}{\memG}{x}{\memD}{\memG}}
%
\inferrule[abs-lam]{
    \tdeabs{\Phi}{\tc}{\qity}{\memD}{\ity}{\memD'}\\
    \edeabs{\tc}{\es;\Upsilon;\Phi}{\qitm}{\memD'}{\memG}{\itm}{\memD''}{\memG'}
}{
    \edeabs{\tc}{\es;\Upsilon;\Phi}{\ilam{\qity}{x}{\qitm}}{\memD}{\memG}{\ilam{\tau}{x}{\itm}}{\memG'}{\memD''}
}
\end{mathpar}
\begin{mathpar}\small
\inferrule[abs-anatrans-stored]{
    n : \st \leadsto \iota/x : \tau \in \memG
}{
    \edeabs{\tc}{\argEnv}{\anatrans{n}{\st}}{\memG}{\memD}{x}{\memG}{\memD}
}

\inferrule[abs-syntrans-stored]{
    n : \st \leadsto \iota/x : \tau \in \memG
}{
    \edeabs{\tc}{\argEnv}{\syntrans{n}}{\memG}{\memD}{x}{\memG}{\memD}
}\end{mathpar}
\begin{mathpar}\small
\inferrule[k-itm-anatrans]{
    n' < n\\
    \sofkX{\st}{\kty}
}{
    \sofkX{\sqitm{\anatrans{n'}{\st}}}{\kitm}
}
\end{mathpar}
\begin{mathpar}\small
\inferrule[abs-syntrans-new]{
    n \notin \text{dom}(\memG)\\
    \keyw{nth}[n](\es) = e\\
    \esynX{e}{\st}{\iota}\\
    \tdeabs{\Phi}{\tc}{\srep{\st}}{\memD}{\ity}{\memD'}\\
    (x~\text{fresh})
}{
    \edeabs{\tc}{\es;\Upsilon;\Phi}{\syntrans{n}}{\memG}{\memD}{x}{\memG, n : \st \leadsto \iota/x : \tau}{\memD'}
}
\end{mathpar}

\begin{mathpar}\small
\inferrule[etctx-emp]{ }{\vdash_\Phi \emptyset \leadsto \emptyset}

\inferrule[etctx-ext]{\vdash_\Phi \Upsilon \leadsto \Gamma\\\istype{\st}{\Phi}\\\vdash_\Phi \st \leadsto \tau}{\vdash_\Phi \Upsilon, x \Rightarrow \st \leadsto \Gamma, x : \tau}
\end{mathpar}

\[\small
\begin{array}{lll}
\textbf{Description}& \textbf{Concrete Form}~~~~ & \textbf{Desugared Form}\\
\text{index projection} & e_\text{targ}\texttt{\#}n & \etarg{\opname{idx}}{n}{e_\text{targ}}{\cdot}\\
\text{label projection} & e_\text{targ}\texttt{\#}\conclbl{lbl} & \etarg{\opname{prj}}{\conclbl{lbl}}{e_\text{targ}}{\cdot}\\
\text{explicit invocation}~~~~ & e_\text{targ}{\cdot}\opname{op}[\sttmidx](\es) & \etarg{\opname{op}}{\sttmidx}{e_\text{targ}}{\es}\\
& e_\text{targ}{\cdot}\opname{op}(\es) & \etarg{\opname{op}}{()}{e_\text{targ}}{\es}\\
& e_\text{targ}{\cdot}\opname{op}(\conclbl{lbl}_1=e_1, \ldots, \conclbl{lbl}_n=e_n) & \keyw{targ}[\opname{op}; \texttt{[}\conclbl{lbl}_1, \ldots, \conclbl{lbl}_n\texttt{]}](e_\text{targ}; \\
& & \quad e_1; \ldots; e_n)\\
\text{labeled case analysis} & e_\text{targ}{\cdot}\opname{case}\,\{ & \keyw{targ}[{\opname{case}}; {\texttt{[}\st_1, \ldots, \st_n\texttt{]}}]({e_\text{targ}}; \\
 & ~|~~\st_1\langle x_1, \ldots, x_k \rangle \Rightarrow e_1 & \quad \eanalam{x_1}{\ldots \eanalam{x_k}{e_1}};\\
 & ~|~~\ldots & \quad \ldots;\\ 
 & ~|~~\st_n\langle x_1, \ldots, x_k \rangle \Rightarrow e_n\} & \quad \eanalam{x_1}{\ldots \eanalam{x_k}{e_n}})
\end{array}
\]

 For example, 
\begin{mathpar}
\small
\inferrule[abs-prod]{
    \tdeabs{\Phi}{\tc}{\qity_1}{\memD}{\ity_1}{\memD'}\\
    \tdeabs{\Phi}{\tc}{\qity_2}{\memD'}{\ity_2}{\memD''}
}{
    \tdeabs{\Phi}{\tc}{\qity_1 \times \qity_2}{\memD}{\ity_1 \times \ity_2}{\memD''}
}
\end{mathpar}

The argument interfaces that populate the list provided to opcon definitions is derived from the argument list by the judgement $\keyw{args}(\es)=_n \stx{args}$, defined as follows:
\begin{mathpar}\small
\inferrule[args-z]{ }{
    \keyw{args}(\cdot)=_0 \skap{\karg}{\svar{nil}}
}

\inferrule[args-s]{
    \keyw{args}(\es)=_{n} \st
}{
    \keyw{args}(\es;e)=_{n+1} \skap{\karg}{\svar{rcons}}~\st~(\slam{\kty}{\svar{ty}}{\sana{n}{\svar{ty}}},\slam{\kunit}{\_}{\ssyn{n}})
}
\end{mathpar}
We assume that the definitions of the standard helper functions $\small\svar{nil} :: \kforall{\kalpha}{\klist{\kalpha}}$ and $\small\svar{rcons} :: \kforall{\kalpha}{\klist{\kalpha} \rightarrow \kalpha \rightarrow \klist{\kalpha}}$, which adds an item to the end of a list, have been substituted into these rules. The result is that the $n$th element of the argument interface list simply wraps the static terms $\sana{n}{\st}$ and $\ssyn{n}$.\end{document}

### 2014-11-01 att-pldi15/att-pldi15.tex
Revised through sec 5

@@
 Only the $\rightarrow$ type constructor (tycon) is built in; all other external tycons (we show ordered record types and constrained string types) are defined by extending a \emph{tycon context}.
-Each defines the semantics of associated term-level operators using functions written in a static language where types and translations are values.
+Each tycon defines the semantics of associated term-level operators using functions written in a static language where types and translations are values.
 The semantics maintain several strong metatheoretic guarantees, notably \emph{type safety} and \emph{conservativity}: that \emph{tycon  invariants} maintained under a minimal tycon context will be conserved under any further extended context.
-Interestingly, mechanized proofs are not needed: problems are caught during typechecking by lifting typed compilation techniques into the semantics and enforcing abstraction barriers around tycons using type abstraction, the same technique that underlies reasoning about ML-style modules.
+Interestingly, mechanized proofs are not needed: problems are caught during typechecking by lifting typed compilation techniques into the semantics and enforcing abstraction barriers around tycons using type abstraction, the same principle that underlies reasoning about ML-style modules.
 \end{abstract}
@@
 In contrast, the usual metatheoretic reasoning techniques for programming languages  (e.g., rule induction) operate on complete language specifications.
-Each {combination} of fragments must formally be treated as its own monolithic language for which  metatheorems must be established anew, guided only informally by those derived for the smaller systems from which it is notionally composed.
+Each {combination} of fragments must formally be treated as its own monolithic language for which  metatheorems must be established anew, guided only informally by those derived for the smaller systems from which the language is notionally composed.
 
@@
 
-These composability questions not arise when a fragment is expressed as an isomorphic embedding (i.e.
+These problems do not arise when a fragment is expressed as an isomorphic embedding (i.e.
 as a library) because modern \emph{module systems} enforce abstraction barriers that ensure that the isomorphism need only be established in the ``closed world'' of the module.
@@
 When library-based embeddings are not possible, as in the examples above, mechanisms are needed that make it possible to define and  reason in a similarly modular manner about  direct extensions to the semantics of a language.
-Such a mechanism could ultimately be integrated directly into the language, blurring the distinction between fragments and libraries and decreasing the need for dialects.% Importing fragments that introduce new semantics   would be as easy and reasonable as importing a new module is in ML today.
+Such a mechanism could ultimately be integrated directly into the language, blurring the distinction between fragments and libraries and decreasing the need for new dialects.% Importing fragments that introduce new semantics   would be as easy and reasonable as importing a new module is in ML today.
 %In the limit, the community could rely on modularly mechanized metatheory and compiler correctness results.% rather than requiring heroic efforts from individual research groups that consider an entire language at once.
@@
 Scala\todo{citation}) because it eliminates the need for type annotations in many situations while remaining decidable in more situations than whole-function type inference and providing what are widely perceived to be higher quality error messages\todo{citation from Dunfield}.
-Secondly, it will give us a clean way to reuse the generalized introductory form, $\eintro{\st}{\es}$, and its associated desugarings at many types, in a manner that relates to recent mechanisms supporting type-specific syntax extensions \cite{TSLs}.
+Secondly, it will give us a clean way to reuse the generalized introductory form, $\eintro{\st}{\es}$, and its associated desugarings, at many types, in a manner that relates to recent mechanisms supporting type-specific syntax extensions \cite{TSLs}.
 For example, when we define regular string types, we will be able to reuse standard string literal syntax.
@@
 We assume the statics of the IL are specified in the standard way by judgements for  type formation {$\iDelta \vdash \tau$}, typing context formation { $\iDelta \vdash \iGamma$} and type assignment {$\iDelta~\iGamma \vdash \iota : \tau\moutput$}.
-The typing and type formation contexts obey standard structural properties (i.e.
-weakening, exchange and contraction).
-In examples, we will omit leading $\emptyset$, used as the base case for finite mappings, and $\cdot$, used as the base case for finite sequences (e.g.
-writing $\Gamma_\text{test} := x : \tau$).
+%The typing and type formation contexts obey standard structural properties (i.e. weakening, exchange and contraction). 
+In examples, we will omit leading $\emptyset$, used as the base case for finite mappings, and $\cdot$, used as the base case for finite sequences (e.g. writing $\Gamma_\text{test} := x : \tau$).
 
@@
 \hspace{-5px}\input{syntax-table-SL}
-\caption{Syntax of the static language (SL). Metavariable $\sx$ ranges over static term variables, $\kalpha$ and $\kvar$ over kind variables and $m$ and $n$ over natural numbers.}
+\caption{Syntax of the static language (SL). Metavariable $\sx$ ranges over static term variables, $\kalpha$ and $\kvar$ over kind variables and $m$ and $n$ over natural numbers.}\vspace{-5px}
 \label{syntax-SL}
@@
 External types, or simply \emph{types}, are static values of kind $\kty$.
-The introductory forms for kind $\kty$ is $\sty{c}{\st}$, where $c$ is a \emph{tycon} and $\st$ is the \emph{type index}.
+The introductory form for kind $\kty$ is $\sty{c}{\st}$, where $c$ is a \emph{tycon} and $\st$ is the \emph{type index}.
 The syntax  for tycons in Fig.
-\ref{syntax-TC} specifies that $c$ is either the built-in tycon governing partial functions, $\rightharpoonup$, a user-defined tycon name written in small caps, $\tc$, or an ``other'' tycon, $\keyw{other}[m; \kappa]$ indexed by a natural number $m$ (to allow arbitrarily many such tycons).
+\ref{syntax-TC} specifies that $c$ is either the built-in tycon governing partial functions, $\rightharpoonup$, a user-defined tycon name written in small caps, $\tc$, or an ``other'' tycon, $\keyw{other}[m]$ indexed by a natural number $m$ (to allow arbitrarily many such tycons).
 The kinding rules governing the form $\sty{c}{\st}$ are shown in Figure \ref{fig:types}.
@@
 we could also have used a list).
-We assume $\small\svar{listfold} :: \kforall{\kalpha_1}{\kforall{\kalpha_2}{\klist{\kalpha_1}\rightarrow\kalpha_2\rightarrow(\kalpha_1\rightarrow\kalpha_2\rightarrow\kalpha_2)\rightarrow\kalpha_2}}$ in defining $\theta_\text{lprod} := \keyw{trans}=\stx{lprod/trans}$  where $\stx{lprod/trans}:=$
+We assume $\small\svar{listfold} :: \kforall{\kalpha_1}{\kforall{\kalpha_2}{\klist{\kalpha_1}\rightarrow\kalpha_2\rightarrow(\kalpha_1\rightarrow\kalpha_2\rightarrow\kalpha_2)\rightarrow\kalpha_2}}$ in defining $\theta_\text{lprod} := \keyw{trans}=\stx{lprod/trans}, \omega_\text{lprod}$  where $\stx{lprod/trans}:=$
 \[\small
@@
 $\memD_\text{paper/abs}  := \stx{title} \leftrightarrow \keyw{str}/\alpha_1, \stx{conf} \leftrightarrow \keyw{str}/\alpha_2$.
-
-Each unique type has a unique type variable associated with it, tracked by the \emph{type translation store}, $\memD$, which maps from types, $\st$, to their (concrete) translations, $\tau$, and the type variables, $\alpha$, which appear in their place in the abstract type translation: $\memD ::= \emptyset ~|~ \memD, \st \leftrightarrow \ity/\alpha$.
+The type translation store $\memD ::= \emptyset ~|~ \memD, \st \leftrightarrow \ity/\alpha$ maintains the  mapping from types, $\st$, to their (non-abstract) translations, $\tau$, and the type variables, $\alpha$, which appear in their place.
+
 Each type translation store induces a \emph{type substitution}, $\delta$, and a corresponding internal type formation context, $\Delta$, according to the judgement $\memD \leadsto \delta : \Delta$.
@@
 For example, $[\delta_\text{paper/abs}]\tau_\text{paper/abs}$ is $\tau_\text{paper} := \keyw{str} \times (\keyw{str} \times \iunit)$, i.e.
-the (concrete) translation of $\stx{paper}$.
+the final type translation of $\stx{paper}$.
 Indeed, we can now give the rule defining the type translation judgement, $\vdash_\Phi \st \leadsto \tau$, mentioned in Sec.
@@
 
-The selective type abstraction judgement is defined by recursing generically over sub-terms until terms of the form $\srep{\st}$ are encountered (see supplement).
-
-The translation of partial function types is direct and is not held abstract, so that lambdas can be used as the sole binding construct in the language:
+With this intuition, we can now give the rules for the selective type abstraction judgement.
+We recurse generically over sub-terms of $\qity$ until sub-terms of form $\srep{\st}$ are encountered (see supplement).
+
+The translation of partial function types is direct and is not held abstract, so that lambdas can be used as the sole binding construct in the external language:
 \begin{mathpar}
@@
 Per above, equivalent types must be  syntactically identical at normal form, and we consider analysis only if $\istype{\st}{\Phi}$, so the rule (subsume) is straightforward.
-
 To use an analytic term in a synthetic position, the programmer must provide a type ascription, written $e : \st$.
@@
   \edeabs{\tc}{{\es;\Upsilon;\Phi}}{\qitm}{\emptyset}{\emptyset}{\iota_\text{abs}}{\memD}{\memG}\\
-  \memG \leadsto \gamma : \Gamma\\\\
+  \memG \leadsto \gamma : \Gamma_\text{abs}\\\\
   \tdeabs{\Phi}{\tc}{\srep{\sty{\tc}{\sttyidx}}}{\memD}{\ity_\text{abs}}{\memD'}\\
-  \memD' \leadsto \delta : \Delta\\\\
-  \Delta~\Gamma \vdash \iota_\text{abs} : \tau_\text{abs}
+  \memD' \leadsto \delta : \Delta_\text{abs}\\\\
+  \Delta_\text{abs}~\Gamma_\text{abs} \vdash \iota_\text{abs} : \tau_\text{abs}
 }{
@@
   \edeabs{\tc}{{(e_\text{targ}; \es);\Upsilon;\Phi}}{\qitm}{\emptyset}{\emptyset}{\iota_\text{abs}}{\memD}{\memG}\\
-  \memG \leadsto \gamma : \Gamma\\\\
+  \memG \leadsto \gamma : \Gamma_\text{abs}\\\\
   \tdeabs{\Phi}{\tc}{\srep{\st}}{\memD}{\ity_\text{abs}}{\memD'}\\
-  \memD' \leadsto \delta : \Delta\\\\
-  \Delta~\Gamma \vdash \iota_\text{abs} : \tau_\text{abs}
+  \memD' \leadsto \delta : \Delta_\text{abs}\\\\
+  \Delta_\text{abs}~\Gamma_\text{abs} \vdash \iota_\text{abs} : \tau_\text{abs}
 }{
@@
     \stx{def}~\stx{args} \Downarrow_{\es; \Upsilon; \Phi} \sqitm{\qitm}\\\\
-    \edeabs{\keyw{other}[m]}{\es; \Upsilon;\Phi}{\qitm}{\emptyset}{\emptyset}{\itm_\text{abs}}{\memD}{\memG}\\
-    \memG \leadsto \gamma : \Gamma\\\\
-    \tdeabs{\Phi}{\keyw{other}[m]}{\srep{\sty{\keyw{other}[m]}{\sttyidx}}}{\memD}{\tau_\text{abs}}{\memD'}\\
-    \memD' \leadsto \delta : \Delta\\\\
-    \Delta~\Gamma \vdash \iota_\text{abs} : \tau_\text{abs}
+    \edeabs{\keyw{other}[m]}{\es; \Upsilon;\Phi}{\qitm}{\emptyset}{\emptyset}{\itm_\text{abs}}{\memD}{\memG}\\\\
+    \tdeabs{\Phi}{\keyw{other}[m]}{\srep{\sty{\keyw{other}[m]}{\sttyidx}}}{\memD}{\tau_\text{abs}}{\memD'}\\\\
+    \memD' \leadsto \delta : \Delta_\text{abs}\\
+    \memG \leadsto \gamma : \Gamma_\text{abs}\\
+    \Delta_\text{abs}~\Gamma_\text{abs} \vdash \iota_\text{abs} : \tau_\text{abs}
 }{
@@
 }
-~~~~~~~~
+~~~~~~
 \inferrule[syn-targ-other]{
@@
   %\snorm{\srep{\sty{\tc}{\sttyidx}}}{\memD}{\cdot}{{\rightharpoonup}, \tc;\emptyset;\Phi}{\sqity{\ity_\text{abs}}}{\memD'}{\cdot}\\\\
-  \edeabs{\keyw{other}[m]}{{(e_\text{targ}; \es);\Upsilon;\Phi}}{\qitm}{\emptyset}{\emptyset}{\iota_\text{abs}}{\memD}{\memG}\\
-  \memG \leadsto \gamma : \Gamma\\\\
-  \tdeabs{\Phi}{\keyw{other}[m]}{\srep{\st}}{\memD}{\ity_\text{abs}}{\memD'}\\
-  \memD' \leadsto \delta : \Delta\\\\
-  \Delta~\Gamma \vdash \iota_\text{abs} : \tau_\text{abs}
+  \edeabs{\keyw{other}[m]}{{(e_\text{targ}; \es);\Upsilon;\Phi}}{\qitm}{\emptyset}{\emptyset}{\iota_\text{abs}}{\memD}{\memG}\\\\
+  \tdeabs{\Phi}{\keyw{other}[m]}{\srep{\st}}{\memD}{\ity_\text{abs}}{\memD'}\\\\
+  \memD' \leadsto \delta : \Delta_\text{abs}\\
+  \memG \leadsto \gamma : \Gamma_\text{abs}\\
+  \Delta_\text{abs}~\Gamma_\text{abs} \vdash \iota_\text{abs} : \tau_\text{abs}
 }{
      \esynX{\etarg{\opname{op}}{\stx{def}}{e_\text{targ}}{\es}}{\st}{[\delta][\gamma]\iota_\text{abs}}
-}
+}\vspace{-5px}
 \end{mathpar}
@@
     \vdash_\Phi \tcstructc{\omega}{op}{\stx{def}} \sim \tcsig{\ktyidx}{\chi, \opsig{\opname{op}}{\klitidx}}
-}
-\end{mathpar}\vspace{-5px}
+}\vspace{-5px}
+\end{mathpar}
 \caption{Checking opcon structures against tycon signatures.}
@@
 The term index is the corresponding static value of kind $\keyw{Str}$, indicated by a subscript for clarity.
-Similarly, for labeled products, records, objects and so on, we have a generalized labeled collection form, $\{\mathtt{lbl}_1=e_1, \ldots, \mathtt{lbl}_n=e_n\}$, that desugars to $\eintro{\texttt{[}\mathtt{lbl}_1, \ldots, \mathtt{lbl}_n\texttt{]}}{e_1; \ldots; e_n}$, i.e.
+Similarly, for labeled products, records, objects and so on, we can define a generalized labeled collection form, $\{\mathtt{lbl}_1=e_1, \ldots, \mathtt{lbl}_n=e_n\}$, that desugars to $\eintro{\texttt{[}\mathtt{lbl}_1, \ldots, \mathtt{lbl}_n\texttt{]}}{e_1; \ldots; e_n}$, i.e.
 a list constructed from the row labels is the term index and the corresponding row values are the arguments.
@@
 
-The first premise of (ana-intro) extracts the tycon definition for the tycon of the type the literal is being analyzed against from $\Phi$.
+The first premise of (ana-intro) extracts the tycon definition for the tycon of the type the intro form is being analyzed against.
 In this example, this is $\tcvar{lprod}$.
@@
 
-The second premise extracts the \emph{intro index kind}, $\klitidx$, from the \emph{opcon signature}, $\chi$, of the tycon signature and the third premise checks the provided term index against this kind.
+The second premise extracts the \emph{intro term index kind}, $\klitidx$, from the \emph{opcon signature}, $\chi$, of the tycon signature and the third premise checks the provided term index against this kind.
 This is simply the kind of term index expected by the tycon, e.g., $\tcvar{lprod}$ specifies $\klist{\klbl}$, so that it can use the labeled collection form, while $\tcvar{rstr}$ specifies an intro index kind of $\keyw{Str}$, so that it can use the string literal form.
@@
 The function has access to the type index, the term index and an interface to the list of arguments.
-The  judgement $\vdash_\Phi \omega \sim \psi$, which appeared as the final premise of the rule (tcc-ext),  checks that the intro opcon definition is well-kinded (rule (ocstruct-intro) in Figure \ref{ocstruct}).
+The  judgement $\vdash_\Phi \omega \sim \psi$, which appeared as the final premise of the rule (tcc-ext),  checks that the intro opcon definition is well-kinded: rule (ocstruct-intro) in Figure \ref{ocstruct}.
 
@@
 \end{array}\]
-
 Because regular strings are implemented as strings, this intro opcon definition is straightforward.
-It begins by making sure that no arguments were passed in (we will return to arguments and the fifth and sixth premises of (ana-intro) with the next example), using the helper function $\svar{arity0} :: \kargs \rightarrow \kunit$ defined such that any non-empty input will raise an error, via the static term $\sraise{\kunit}$.
+It begins by making sure that no arguments were passed in (we will return to arguments and the fifth and sixth premises of (ana-intro) with the next example), using the helper function $\svar{arity0} :: \kargs \rightarrow \kunit$ defined such that any non-empty list will raise an error, via the static term $\sraise{\kunit}$.
 In practice, the tycon provider would specify an error message here.%The kinding rules and operational semantics of this exception-like form are standard:
@@
 \ref{sec:type-translations}.
-Also like translational internal types, there is an unquote form, $\quq{\st}$, that permits translational internal terms to be constructed compositionally.
-%The unquote form is eliminated in the course of evaluation.
-The supplement gives the complete rules.
+Also analagously, there is an unquote form, $\quq{\st}$, that permits translational internal terms to be constructed compositionally.
+The supplement gives the analagous rules.
 
 The final two forms of translational internal term are $\anatrans{n}{\st}$ and $\syntrans{n}$.
-These represent the translation of argument $n$, the first if it arises via analysis against type $\st$ and the second if it arises by synthesis.
+These stand in for the translation of argument $n$, the first if it arises via analysis against type $\st$ and the second if it arises via type synthesis.
 They are not intended to be written directly by tycon providers, arising only internally in the semantics of argument interface lists.
-Before giving the rules, let us motivate the mechanism with the intro opcon definition for $\tcvar{lprod}$, $\omega_\text{lprod} := \tcstructn{\stx{lprod/intro}}; \omega_\text{lprod/targops}$ where $\stx{lprod/intro} :=$
+Before giving the rules, let us motivate the mechanism with the intro opcon definition for $\tcvar{lprod}$.
+We have $\omega_\text{lprod} := \tcstructn{\stx{lprod/intro}}; \omega_\text{lprod/targops}$ where $\stx{lprod/intro} :=$
 \[\small
@@
 \]
-
 The first line checks that the type provided is inhabited, in this case by checking that there are no duplicate labels via the helper function $\svar{uniqmap} :: \klist{\klbl \times \kty}\rightarrow \kunit$, raising an error if there are.
 An alternative strategy may have been to use an abstract kind that ensured that such type indices could not have been constructed, but to be compatible with our equality kind restriction, this would require support for abstract equality kinds, analagous to abstract equality types in SML.
-We chose not to formalize these for simplicity, and to demonstrate this more general technique: types with indices that are to be considered ``malformed'' can simply be left uninhabited.
+We chose not to formalize these for simplicity, and to demonstrate this general technique.
 An analagous technique could be used to implement record types by requiring that the index be sorted.%, leaving indices where the labels did not appear sorted uninhabited.
@@
 
-\paragraph{Argument Interface Lists} We define $\karg$, the kind of \emph{argument interfaces}, as a simple product of functions, $\karg := (\kty \rightarrow \kitm) \times (\kunit \rightarrow \kty\times\kitm)$, one for analysis and the other for synthesis, both having the expected type.
+\paragraph{Argument Interface Lists} We define $\karg$, the kind of \emph{argument interfaces}, as a simple product of functions, $\karg := (\kty \rightarrow \kitm) \times (\kunit \rightarrow \kty\times\kitm)$, one for analysis and the other for synthesis, both having the expected kind for these operations.
 The helper functions $\svar{ana}$ and $\svar{syn}$ simply project the corresponding function out, $\svar{ana} := \slam{\karg}{\svar{arg}}{\keyw{fst}(\svar{arg})}$ and $\svar{syn}  :=  \slam{\karg}{\svar{arg}}{\keyw{snd}(\svar{arg})}$.
 
-The argument interface list, called $\stx{args}$ in the rules, is simply a static list where the $i$th entry is a pair of functions of the form $(\slam{\kty}{\svar{ty}}{\sana{i}{\svar{ty}}}, \slam{\kunit}{\_}{\ssyn{i}})$.
-It is generated by the judgement $\keyw{args}(n)=\stx{args}$, where $n$ is given as the length of the argument list, written $|\es|=n$.
-See supplement.
+The argument interface list, called $\stx{args}$ in the rules, is simply a static list of kind $\kargs$ where the $i$th entry is $(\slam{\kty}{\svar{ty}}{\sana{i}{\svar{ty}}}, \slam{\kunit}{\_}{\ssyn{i}})$.
+It is generated by the judgement $\keyw{args}(n)=\stx{args}$, where $n$ is the length of the argument list, written $|\es|=n$ (see supplement).
 
@@
 
- Recall that the kinding judgement is indexed by $n$, which is an upper bound on the argument index of such terms. This is enforced in the kinding rules:
+ Recall that the kinding judgement is indexed by $n$, which is an upper bound on the argument index of terms of the form $\sana{n}{\st}$ and $\ssyn{n}$. This is enforced in the kinding rules:
 \begin{mathpar}\small
@@
 
-The following lemma thus characterizes the argument interface list judgement:
+The following lemma characterizes thus characterizes the argument list interface generation judgement:
 \begin{lemma}
@@
 The rule (ocstruct-intro) ruled out writing either of these forms explicitly in an opcon definition by checking against the bound $n=0$.
-This is to prevent out-of-bounds errors -- tycon providers do not write these forms directly, but only access them via the argument interface list constructed by the semantics.
+This is to prevent out-of-bounds errors -- tycon providers do not write these forms directly, but only access them via the argument interface list, guaranteed to have the correct length.
 This is possible because:
@@
 We write $\small[\Upsilon \vdash_\Phi e \nLeftarrow \st]$ to indicate that $e$ fails to analyze against $\st$.
-Because we do not directly define this judgement  inductively, we also allow that this premise be left out, leaving a non-deterministic semantics, which does not affect our metatheory.
+We do not define this  inductively, so we also allow that this premise be omitted, leaving a non-deterministic semantics nevertheless sufficient for our metatheory.
 
@@
 
-The kinding rules also prevent the translational internal term forms generated by these static operations from being well-kinded when $n = 0$ so they can only arise in this manner (see supplement).
+The kinding rules also prevent these translational internal term forms generated by these static operations from being well-kinded when $n = 0$.
 Like the translational internal type form $\srep{\st}$, these  are retained in values of kind $\kitm$:
@@
 \paragraph{Abstracted Term Translations} The reason that the translations themselves were not inserted yet (even though they were derived in rule (s-ana-success) above) is again because we want to be able to  hold these abstract by replacing them with variables.
-The judgement $\edeabs{\tc}{\argEnv}{\qitm}{\memD}{\memG}{\itm}{\memD'}{\memG'}$, appearing as the eighth premise of (ana-intro), relates a translational internal term $\qitm$ to an internal term $\itm$ called the corresponding \emph{abstracted term translation}, where all references to the translation of an argument (having any type) are replaced with a fresh variable and all references to the translation of a type constructed by a user-defined tycon other than the ``delegated tycon'' $\tc$ are replaced with an abstract type variable as described in Sec.
+The judgement $\edeabs{\tc}{\argEnv}{\qitm}{\memD}{\memG}{\itm}{\memD'}{\memG'}$, appearing as the eighth premise of (ana-intro), relates a translational internal term $\qitm$ to an internal term $\itm$ called the corresponding \emph{abstracted term translation}, where all references to the translation of an argument (having any type) are replaced with a unique variable and all references to the translation of a type constructed by a user-defined tycon other than the ``delegated tycon'' $\tc$ are replaced with an abstract type variable as described in Sec.
 \ref{sec:selective-type-translation-abstraction}.
@@
 
-Each $\memG$ induces an internal term substitution,  $\gamma ::= \emptyset ~|~ \gamma, \iota/x$, and corresponding internal typing context $\Gamma$ by the judgement $\memG \leadsto \gamma : \Gamma$, used as the ninth premise.
+Like $\memD$, each $\memG$ induces an internal term substitution,  $\gamma ::= \emptyset ~|~ \gamma, \iota/x$, and corresponding internal typing context $\Gamma$ by the judgement $\memG \leadsto \gamma : \Gamma$, appearing as the ninth premise.
 In this case, $\small
-\gamma_\text{ex/abs} := title/x_0, \texttt{"EXMPL 2015"}/x_1$ and $\small\Gamma_\text{ex/abs} := x_0 : \alpha_0, x_1 : \alpha_1$.
+\gamma_\text{ex/abs} := title/x_0, \texttt{"EXMPL 2015"}_\text{IL}/x_1$ and $\small\Gamma_\text{ex/abs} := x_0 : \alpha_0, x_1 : \alpha_1$.
 
@@
 Nevertheless, because our intro opcon definition treated arguments parametrically, the check succeeds.
-We will described  an ill-behaved operator, which would lead to a violation of the regular string tycon  invariant were it not disallowed by this check,  in Sec.
+We will describe  an ill-behaved operator, which would lead to a violation of the regular string tycon  invariant were it not disallowed by this check,  in Sec.
 \ref{metatheory}.
@@
 Applying the substitutions $\gamma_\text{ex/abs}$ and $\delta_\text{ex/abs}$ in the conclusion of the rule, we arrive at $\iota_\text{ex}$.
+
+The rule (ana-intro-other) is used to introduce values of a type constructed by an ``other'' tycon.
+The term index, rather than the tycon context, directly specifies the static function that maps the arguments to a translation.
+In all other respects, the rule is analagous.
+It is used as a technical device in our proof of conservativity in Sec.
+\ref{metatheory}.
 
@@
 All non-introductory opcons associated with user-defined tycons go through another generalized form, in this case for \emph{targeted operations}, $\etarg{\opname{op}}{\sttmidx}{e_\text{targ}}{\es}$, where $\opname{op}$ ranges over opcon names, ${\sttmidx}$ is the term index, $e_\text{targ}$ is the \emph{target argument} and $\es$ are the remaining arguments.
-Some examples of desugarings to this form include explicit invocation, $e_\text{targ}.\opname{op}\langle\sttmidx\rangle(\es)$, variants where the term index or the arguments are omitted, and label projection syntax for use by labeled product-like types, $e_\text{targ}\opname{\#}\conclbl{lbl}$, which desugars to $\etarg{\opname{\#}}{\conclbl{lbl}}{e_\text{targ}}{\cdot}$, and concatenative syntax $e_\text{targ} \cdot e_\text{arg}$, which desugars to $\etarg{\opname{conc}}{\striv}{e_\text{targ}}{e_\text{arg}}$.
-We show others, including case analysis, in the supplement.
-
-Whereas introductory operations were strictly analytic, targeted operations are strictly synthetic in @$\lambda$.
-The type and translation is determined by the tycon of the type synthesized by the target argument (i.e.
-this tycon is used as the delegated tycon).
+Some examples of desugarings to this form include explicit invocation, $e_\text{targ}.\opname{op}\langle\sttmidx\rangle(\es)$ (and variants of where the term index or the arguments are omitted), projection syntax for use by record-like types, $e_\text{targ}\opname{\#}\conclbl{lbl}$, which desugars to $\etarg{\opname{\#}}{\conclbl{lbl}}{e_\text{targ}}{\cdot}$, and concatenative syntax $e_\text{targ} \cdot e_\text{arg}$, which desugars to $\etarg{\opname{conc}}{\striv}{e_\text{targ}}{e_\text{arg}}$.
+We show other desugarings, including case analysis, in the supplement.
+
+Whereas introductory operations were analytic, targeted operations are synthetic in @$\lambda$.
+The type and translation is determined by the tycon of the type synthesized by the target argument.
 The rule (syn-targ) is otherwise similar to (ana-intro) in its structure.
@@
 The third extracts the \emph{operator index kind} from its opcon signature, and the fourth checks the term index against this kind.
-For example, we may associate the following targeted opcon signatures with the tycons $\tcvar{rstr}$ and $\tcvar{lprod}$:\[\small
+For example, we may associate the following opcon signatures with the tycons $\tcvar{rstr}$ and $\tcvar{lprod}$:\[\small
 \begin{array}{lcl}
-\chi_\text{rstr/targops} & := & \opsig{\opname{conc}}{\kunit}, \opsig{\opname{case}}{\klist{\keyw{StrPattern}}}, \opsig{\opname{coerce}}{\krx}, \\
-& & \opsig{\opname{check}}{\krx}, \opsig{\opname{replace}}{\krx}\\
-\chi_\text{lprod/targops} & := & \opsig{\opname{prj}}{\klbl}, \opsig{\opname{conc}}{\klist{\klbl}}, \opsig{\opname{drop}}{\klist{\klbl}}
+\chi_\text{rstr} & := & \keyw{intro}[\keyw{Str}], \opsig{\opname{conc}}{\kunit}, \opsig{\opname{case}}{\klist{\keyw{StrPattern}}},  \\
+& & \opsig{\opname{coerce}}{\krx}, \opsig{\opname{check}}{\krx}, \opsig{\opname{replace}}{\krx}\\
+\chi_\text{lprod} & := & \keyw{intro}[\klist{\klbl}], \opsig{\opname{prj}}{\klbl}, \opsig{\opname{conc}}{\kunit}, \opsig{\opname{drop}}{\klist{\klbl}}
 \end{array}
@@
 The opcons associated with $\tcvar{rstr}$ are taken directly from Fulton et al.'s specification of regular string types \cite{sanitation-psp14}, with the exception of $\opname{case}$, which generalizes case analysis as defined there to arbitrary string patterns, which we discuss in the supplement.
-The opcons associated with $\tcvar{lprod}$ are also straightforward: $\opname{prj}$ projects out a labeled value from a labeled product, $\opname{conc}$ creates a new labeled product based on the target with some fields updated or added from the argument, and $\opname{drop}$ creates a new labeled product from the target with some fields dropped.
-Because targeted operations use no new mechanisms, we will only show regular string concatenation here.
-The others are in the supplement.
+The opcons associated with $\tcvar{lprod}$ are also straightforward: $\opname{prj}$ projects out the row with the provided label, $\opname{conc}$ concatenates two labeled products (updating common fields with the value of the right argument), and $\opname{drop}$ creates a new labeled product from the target with some fields dropped.
+Note that both regular strings and labeled products define concatenation.
+Targeted operations use no new mechanisms, so we will only show regular string concatenation here; others are in the supplement.
 
@@
 For example, $\omega_\text{rstr/targops}$ defines:
-\vspace{-3px}
 \[\small
@@
 \end{array}
-\]\vspace{-3px}
+\]
 The first line simply checks that exactly two total arguments, including the target term, were passed in using the helper function $\svar{arity2}$.
@@
 
-The remaining premises are identical to the corresponding premises in (ana-intro), with the only exception being that we check the abstract term translation against the abstract type translation of the synthesized type.
-In this example, only regular string types are involved so the abstracted type translations will not differ from the type translations themselves, i.e.
-$\keyw{str}$ (though abstracted arguments are still used to ensure hygiene, see Sec.
-\ref{sec:hygiene}).
-
-\subsection{Other} The final form of external term is $\eother{\st}{\es}$.
-Rule (ana-other) specifies that it can be analyzed against $\sotherty{m}{\tau}$ if $\iota$ has type $\tau$ in the translation of the current typing context.
-We will see how it is used technically in the next section.
-\begin{mathpar}\small
-
-\end{mathpar}
+The remaining premises of (syn-targ) are analagous to the corresponding premises in (ana-intro), with the only exception being that we check the abstract term translation against the abstract type translation of the synthesized type.
+Like (ana-intro-other), the rule (syn-targ-other) is used when the target synthesizes an ``other'' type.
+The mapping from the arguments to a type and translation is given directly in the term index (the operator name is ignored).
 
@@
 \section{Metatheory}\label{metatheory}
-We will now state and sketch proofs of the key metatheoretic properties of @$\lambda$. The full proofs are in the supplement.
-
+We will now state the key metatheoretic properties of @$\lambda$. The full proofs are in the supplement.
+
+\paragraph{Kind Safety} Kind safety ensures that normalization of well-kinded static terms cannot go wrong. We can take a standard progress and preservation based approach. 
+\begin{theorem}[Static Progress]
+If $\sofkn{\emptyset}{\emptyset}{\Phi}{n}{\st}{\kappa}$ and $\vdash \Phi$ and $|\es|=n$ and $\vdash_\Phi \Upsilon \leadsto \Gamma$ then $\sval{\st}{\es; \Phi; \Upsilon}$ or $\serr{\st}{\es; \Phi; \Upsilon}$ or $\sstep{\st}{\es; \Phi; \Upsilon}{\st'}$.
+\end{theorem}
+
+\begin{theorem}[Static Preservation]
+If $\sofkn{\emptyset}{\emptyset}{\Phi}{n}{\st}{\kappa}$ and $\vdash \Phi$ and $|\es|=n$ and $\vdash_\Phi \Upsilon \leadsto \Gamma$ and $\sstep{\st}{\es; \Phi; \Upsilon}{\st'}$ then $\sofkn{\emptyset}{\emptyset}{\Phi}{n}{\st'}{\kappa}$.
+\end{theorem}
+
+The case for the form $\ssyn{n}$ requires that the following theorem be simultaneously defined.
+The mutual induction is well-founded because the total length of all the argument lists in the terms being considered monotonically decreases.
+\begin{theorem}[Type Synthesis]
+If $\vdash \Phi$ and $\vdash_\Phi \Upsilon \leadsto \Gamma$ and $\esynX{e}{\st}{\iota}$ then $\vdash_\Phi \st \leadsto \tau$ (and thus $\istype{\st}{\Phi}$). 
+\end{theorem}
 \paragraph{Type Safety}
-Type safety requires that every well-typed external term translate to a well-typed internal term.
-Combined with type safety for the IL \cite{pfpl}, we have that well-typed external terms cannot go wrong.
-To prove this theorem, we must actually prove a stronger theorem, type-preserving translation, the analog of the notion of type-preserving compilation in typed compilation \cite{tarditi+:til-OLD}\todo{right citation for TIL?}.% produced as the translation of the external type.
+Type safety in a typed translation semantics requires that well-typed external terms translate to well-typed internal terms.
+Type safety for the IL \cite{pfpl} then implies that well-typed external terms cannot go wrong.
+To prove this, we must prove a stronger theorem, type-preserving translation (the analog of type-preserving compilation in the typed compilation literature \cite{tarditi+:til-OLD}\todo{right citation for TIL?}).% produced as the translation of the external type.
 Note that we need only to state our top-level theorems for the analysis judgement  because of the subsumption rule.
@@
 \begin{theorem}[Type Preserving Translation]
-If $\vdash \Phi$ and $\vdash_\Phi \Upsilon \leadsto \Gamma$ and $\istype{\st}{\Phi}$ and $\vdash_\Phi \st \leadsto \tau$ and $\eanaX{e}{\st}{\iota}$ then $\emptyset~\Gamma \vdash \iota : \tau$.
+If $\vdash \Phi$ and $\vdash_\Phi \Upsilon \leadsto \Gamma$ and $\vdash_\Phi \st \leadsto \tau$ and $\eanaX{e}{\st}{\iota}$ then $\emptyset~\Gamma \vdash \iota : \tau$.
 \end{theorem}
 \begin{proof-sketch}
-We proceed by induction on the typing judgement.
-The interesting cases are (ana-intro) and (syn-trans).
-The result follows directly from the final premise of these rules, combined with a lemma that states that the substitutions $\gamma$ and $\delta$ generated by the abstracted type and term translation stores satisfy $\Delta$ and $\Gamma$, and thus that applying them results in a well-typed term.
+We induct on the typing judgement.
+The interesting cases are (ana-intro), (ana-intro-other), (syn-trans) and (syn-trans-other); the later two arise via subsumption.
+The result follows directly from the final premise of each rule, combined with a lemma that states that the substitutions $\gamma$ and $\delta$ generated by the abstracted type and term translation stores satisfy the corresponding $\Delta_\text{abs}$ and $\Gamma_\text{abs}$, and so applying them in the conclusion gives a well-typed term.
 \end{proof-sketch}
@@
 \paragraph{Hygienic Translation} 
-Note in rules (ana-intro) and (syn-trans) that the domains of $\Upsilon$ and $\Gamma$ are disjoint.
+Note in the just mentioned rules that the domains of $\Upsilon$ and $\Gamma_\text{abs}$ are disjoint.
 This serves to ensure \emph{hygienic translation} -- translations cannot refer to variables in the surrounding scope directly, so uniformly renaming a variable cannot change the meaning of a program.
-Variables in $\Upsilon$ can (and do -- $title$ in the previous example) occur in arguments to the operation, but the translations of the arguments only appear \emph{after} the substitution $\gamma$ has been applied.
-We assume that applying a substitution is capture-avoiding in the usual manner.
+Variables in $\Upsilon$ can (e.g.
+$title$ in the previous example) occur in arguments, but the translations of the arguments only appear \emph{after} the substitution $\gamma$ has been applied.
+We assume that substitution is capture-avoiding in the usual manner.
 %(i.e.
@@
 \paragraph{Unicity}
-The rules are structured so that if a term is well-typed, both its type and translation are unique. This can be stated as a unicity theorem.
+The rules are structured so that if a term is well-typed, both its type and translation are unique.\todo{could move this whole thing to supplement if room needed}
 \begin{theorem}[Unicity]
-If $\vdash \Phi$ and $\vdash_\Phi \Upsilon \leadsto \Gamma$ and $\istype{\st}{\Phi}$ and $\istype{\st'}{\Phi}$ and $\eanaX{e}{\st}{\iota}$ and $\eanaX{e}{\st'}{\iota'}$ then $\st = \st'$ and $\iota = \iota'$.
+If $\vdash \Phi$ and $\vdash_\Phi \Upsilon \leadsto \Gamma$ and $\vdash_\Phi \st \leadsto \tau$ and $\vdash_\Phi \st' \leadsto \tau'$ and $\eanaX{e}{\st}{\iota}$ and $\eanaX{e}{\st'}{\iota'}$ then $\st = \st'$ and $\iota = \iota'$.
 \end{theorem}
@@
 \begin{theorem}[Stability]
-Letting $\Phi' := \Phi, \tcdef{\tc}{\psi}{\theta}$, if $\vdash \Phi'$ and $\vdash_\Phi \Upsilon \leadsto \Gamma$ and $\istype{\st}{\Phi}$ and $\eanaX{e}{\st}{\iota}$ and $\vdash_\Phi \st \leadsto \tau$ then, $\vdash_{\Phi'} \Upsilon \leadsto \Gamma$ and $\istype{\st}{\Phi'}$ and $\eana{\Upsilon}{\Phi'}{e}{\st}{\iota}$ and $\vdash_{\Phi'} \st \leadsto \tau$.
+Letting $\Phi' := \Phi, \tcdef{\tc}{\psi}{\theta}$, if $\vdash \Phi'$ and $\vdash_\Phi \Upsilon \leadsto \Gamma$ and $\vdash_\Phi \st \leadsto \tau$ and $\eanaX{e}{\st}{\iota}$ then, $\vdash_{\Phi'} \Upsilon \leadsto \Gamma$ and $\vdash_{\Phi'} \st \leadsto \tau$ and $\eana{\Upsilon}{\Phi'}{e}{\st}{\iota}$.
 \end{theorem}
@@
 
-\begin{tyconinvariant}[Regular Strings]
-For all $e$ and $\sttyidx=\concrx{r}$, if $\eana{\emptyset}{\Phi_\text{rstr}}{e}{\sty{\tcvar{rstr}}{\sttyidx}}{\iota}$ and $\iota \Downarrow \iota'$ then $\iota'=\texttt{"s"}$ and $\texttt{"s"}$ is in the regular language $\mathcal{L}(\texttt{r})$.
+\begin{tyconinvariant}[Regular String Soundness]
+If $\eana{\emptyset}{\Phi_\text{rstr}}{e}{\sty{\tcvar{rstr}}{\concrx{r}}}{\iota}$ and $\iota \Downarrow \iota'$ then $\iota'=\texttt{"s"}$ and $\texttt{"s"}$ is in the regular language $\mathcal{L}(\texttt{r})$.
 \end{tyconinvariant}
-\begin{proof-sketch} The proof of this invariant is not unusually difficult because we have fixed the tycon context $\Phi_\text{rstr}$, so we can simply reason about the language like a type-directed compiler specification for a calculus with only two tycons, $\rightharpoonup$ and $\tcvar{rstr}$.
+\begin{proof-sketch} The proof is not unusually difficult because we have fixed the tycon context $\Phi_\text{rstr}$, so we can simply treat the calculus like a type-directed compiler for a calculus with only two tycons, $\rightharpoonup$ and $\tcvar{rstr}$.
 Such a calculus and compiler specification was given by Fulton et al.
-\cite{sanitation-psp14} and extending this to our  setting requires showing only that the opcon definitions in $\tcvar{rstr}$ adequately capture these specifications using standard program correctness techniques for the SL.\end{proof-sketch}
-
-When we extend the tycon context, however, it is not trivially obvious that this invariant continues to hold.
-For example, a new tycon might define a targeted operator that synthesizes a regular string type, e.g.
-$\stx{paper}$, and generates a translation that is not in the corresponding regular language, e.g.
-$\texttt{""}$.
-The reason this cannot occur is because we hold translations of regular string types abstract  when passing through an opcon defined by any other tycon, and so, by standard parametricity properties for the IL, and the empty string does not have type $\alpha$.
-Indeed, no string other than one that came through an opcon associated with $\tcvar{rstr}$ can arise, so inductively, we know the invariant is maintained.
-We can state this more generally as a conservativity theorem.
+\cite{sanitation-psp14} and extending this to our  setting requires showing only that the opcon definitions in $\tcvar{rstr}$ adequately capture these specifications using standard program correctness techniques for the SL, which is a simple functional language.
+The rule (syn-targ-other) can synthesize a regular string type, but the abstracted translation will always be checked against $\tau_\text{abs}=\alpha$, so no strings that could not arise by another rule can be generated.
+\end{proof-sketch}
+
+The reason why the rule (syn-targ-other) is not a problem in proving a tycon invariant turns out to be the same reason extending the tycon context also cannot violate such a tycon invariant.
+Any newly introduced tycon defining a targeted operator that synthesizes a regular string type, e.g.
+$\stx{paper}$, and generating a translation that is not in the corresponding regular language, e.g.
+$\texttt{""}$ could be defined, but when used, the rule (syn-targ) would check this translation against $\tau_\text{abs}=\alpha$ and the check would fail.
+%Type abstraction is the basis for conservatively composing type system fragments, just like it is the basis for composing ML-style modules.
+ We can state this conservativity property more generally.
 
 \begin{theorem}[Conservativity] If $\vdash \Phi$ and $\tc \in \text{dom}(\Phi)$ and a tycon invariant for $\tc$ holds under $\Phi$: \begin{itemize}
-\item for all $e$ and $\sttyidx$, if $\eana{\emptyset}{\Phi}{e}{\sty{\tc}{\sttyidx}}{\iota}$ and $\vdash_\Phi {\sty{\tc}{\sttyidx}} \leadsto \tau$ and $\iota \Downarrow \iota'$ then $P(\sttyidx, \iota')$
+\item If $\eana{\emptyset}{\Phi}{e}{\sty{\tc}{\sttyidx}}{\iota}$ and $\vdash_\Phi {\sty{\tc}{\sttyidx}} \leadsto \tau$ and $\iota \Downarrow \iota'$ then $P(\sttyidx, \iota')$.
 \end{itemize} then for all $\Phi' = \Phi, \tcdef{\tc'}{\psi'}{\theta'}$ such that $\vdash \Phi'$, the same tycon invariant holds under $\Phi'$: \begin{itemize}
-\item for all $e$ and $\sttyidx$, if $\eana{\emptyset}{\Phi'}{e}{\sty{\tc}{\sttyidx}}{\iota}$ and $\vdash_{\Phi'}{\sty{\tc}{\sttyidx}}\leadsto{\tau}$ and $\iota \Downarrow \iota'$ then $P(\sttyidx, \iota)$.
+\item If $\eana{\emptyset}{\Phi'}{e}{\sty{\tc}{\sttyidx}}{\iota}$ and $\vdash_{\Phi'}{\sty{\tc}{\sttyidx}}\leadsto{\tau}$ and $\iota \Downarrow \iota'$ then $P(\sttyidx, \iota)$.
 \end{itemize}
@@
 \begin{proof-sketch}
-For (ana-intro), we simply construct an equivalent introductory term that is well-typed in $\Phi$, where arguments that were of a type constructed by $\tc'$ are replaced by $\eother{\iota}$, making use of the fact that type $\sotherty{m}{\tau}$ is indistinguishable from a type outside $\Phi$ to the intro opcon definition for $\tc$ (see supplement for the full definition of this construction).
-For (syn-targ), we use the parametricity property described above if the synthesized type is constructed by $\tc$ to get our result by induction.
-\end{proof-sketch}
+The proof maps every well-typed term in the extended tycon context to a well-typed term in the original tycon context with the same translation, so we can apply the original invariant to arrive at the property $P(\sttyidx, \iota')$.
+This term is constructed by replacing all types constructed by $\tc'$ with a type constructed by $\keyw{other}[m]$, for an $m$ uniquely associated with $\tc'$, and all introductory and targeted sub-terms associated with $\tc'$ with an equivalent one that passes through the rules (ana-intro-other) and (syn-targ-other).\end{proof-sketch}
 %Had the translational term generated by the intro opcon definition been, e.g., $(\anatrans{0}{\stx{title}}, (\texttt{"TEST"}, ())$, then the corresponding abstract term would be $(x_0, (\texttt{"TEST"}, ())$ and the check would fail because $(\alpha_0, \alpha_1)~(x_0 : \alpha_0) \nvdash \texttt{"TEST"} : \alpha_1$.
@@
 
-\paragraph{Optimization and Extensible Compilers}
+\paragraph{Optimization}
 When a simple embedding that preserves the static semantics exists, but a different  embedding would better preserve a desired cost semantics, term rewriting techniques can also be used to perform ``optimizations'', thus achieving an isomorphic embedding.

### 2014-11-01 att-pldi15/att-pldi15.tex
More revisions

@@
 In this paper, we begin from first principles with a core calculus, @$\lambda$, specified like many full-scale languages: as a bidirectionally typed translation semantics.
-Only the $\rightarrow$ type constructor (tycon) is built in; all other external tycons (we show ordered record types and constrained string types) are defined by extending a \emph{tycon context}.
-Each tycon defines the semantics of associated term-level operators using functions written in a static language where types and translations are values.
-The semantics maintain several strong metatheoretic guarantees, notably \emph{type safety} and \emph{conservativity}: that \emph{tycon  invariants} maintained under a minimal tycon context will be conserved under any further extended context.
-Interestingly, mechanized proofs are not needed: problems are caught during typechecking by lifting typed compilation techniques into the semantics and enforcing abstraction barriers around tycons using type abstraction, the same principle that underlies reasoning about ML-style modules.
+Only the $\rightarrow$ type constructor (tycon) is built in; all other external tycons (we show a variant of record types and constrained string types) are defined by extending a \emph{tycon context}.
+Each tycon defines the semantics of associated term-level operators (e.g.
+record projection) using functions written in a static language where types and translations are values.
+The semantics provide  strong metatheoretic guarantees, notably \emph{type safety} and \emph{conservativity}: that all \emph{tycon  invariants} will be conserved under extension.
+Mechanized proofs are not needed: problems are caught during typechecking by lifting typed compilation techniques into the semantics and enforcing abstraction barriers around tycons using type abstraction, the same principle that underlies ML-style modules.
 \end{abstract}
@@
 for data parallelism  \cite{chakravarty2007data}, distributed programming \cite{Murphy:2007:TDP:1793574.1793585}, reactive programming \cite{mandel2005reactiveml}, authenticated data structures \cite{Miller:2014:ADS:2535838.2535851}, databases \cite{Ohori:2011:MSM:2034773.2034815},  units of measure \cite{conf/cefp/Kennedy09} and regular string sanitation \cite{sanitation-psp14}.
-\verb|sprintf| in the OCaml dialect statically distinguishes format strings from strings.% All of these are implemented as dialects of existing languages, presumably because a strong encoding was not feasible.
+%\verb|sprintf| in the OCaml dialect statically distinguishes format strings from strings.% All of these are implemented as dialects of existing languages, presumably because a strong encoding was not feasible.
 
@@
 The introductory form for kind $\kty$ is $\sty{c}{\st}$, where $c$ is a \emph{tycon} and $\st$ is the \emph{type index}.
-The syntax  for tycons in Fig.
-\ref{syntax-TC} specifies that $c$ is either the built-in tycon governing partial functions, $\rightharpoonup$, a user-defined tycon name written in small caps, $\tc$, or an ``other'' tycon, $\keyw{other}[m]$ indexed by a natural number $m$ (to allow arbitrarily many such tycons).
+The syntax  for tycons in Figure \ref{syntax-TC} specifies that $c$ is either the built-in tycon governing partial functions, $\rightharpoonup$, a user-defined tycon name written in small caps, $\tc$, or an ``other'' tycon, $\keyw{other}[m]$ indexed by a natural number $m$ (to allow arbitrarily many such tycons).
 The kinding rules governing the form $\sty{c}{\st}$ are shown in Figure \ref{fig:types}.
@@
     \vdash \Phi\\
-    \tc \notin \text{dom}(\Phi)\\\\
-    \psi = (\tcsig{\ktyidx}{\chi})\\
-    \keq{\emptyset}{\ktyidx}\\\\
-    \theta = (\tcstruct{\stx{schema}}{\omega})\\
+    \tc \notin \text{dom}(\Phi)\\
+    \keq{\emptyset}{\ktyidx}\\
     \sofkz{\emptyset}{\emptyset}{\Phi}{\stx{schema}}{\ktyidx \rightarrow \kity}\\\\
-    \vdash_{\Phi, \tcdef{\tc}{\psi}{\theta}} \omega \sim \psi
-}{
-    \vdash \Phi, \tcdef{\tc}{\psi}{\theta}
+    \vdash_{\Phi,  \tcdef{\tc}{\tcsig{\ktyidx}{\chi}}{\tcstruct{\stx{schema}}{\omega}}} \omega \sim \tcsig{\ktyidx}{\chi}
+}{
+    \vdash \Phi, \tcdef{\tc}{\tcsig{\ktyidx}{\chi}}{\tcstruct{\stx{schema}}{\omega}}
 }\vspace{-8px}
 \end{mathpar}
-\caption{Tycon context well-definedness. We omit the empty case (tcc-emp) for concision.}
+\caption{Tycon context well-definedness. We omit the trivial case for when $\Phi=\cdot$ for concision.}
 \label{fig:tycon-ctxs}
@@
 \section{Typing and Translation of External Terms}\label{external-terms}
+
 Having established how types are constructed, and how they determine abstract and from there concrete translations to internal types, we can finally give the typing and translation rules for external terms, shown in Figure \ref{typing}.
@@
 \cite{pfpl}), and define both only in analytic positions for simplicity.
+
+\subsection{Generalized Introductory Operations}\label{sec:introop}
+The translation of the generalized introductory form, $\eintro{\sttmidx}{\es}$, is determined by the tycon of the type it is being analyzed against as a function of the type's index, the \emph{term index}, $\sttmidx$, and the \emph{argument list}, $\es$.
+
+Before discussing rules (ana-intro) and (ana-intro-other), we note that we can recover a variety of standard concrete introductory forms by a purely syntactic desugaring to this abstract form (and thus allow their use at more than one type).
+For example, for regular strings we will be able to use the string literal form, $\texttt{"s"}$, which desugars to $\eintro{\texttt{"s"}_\text{SL}}{\cdot}$.
+The term index is the corresponding static value of kind $\keyw{Str}$, indicated by a subscript for clarity.
+Similarly, for labeled products, records, objects and so on, we can define a generalized labeled collection form, $\{\mathtt{lbl}_1=e_1, \ldots, \mathtt{lbl}_n=e_n\}$, that desugars to $\eintro{\texttt{[}\mathtt{lbl}_1, \ldots, \mathtt{lbl}_n\texttt{]}}{e_1; \ldots; e_n}$, i.e.
+a list constructed from the row labels is the term index and the corresponding row values are the arguments.
+In both cases, the term index captures  static portions of the concrete form and the arguments capture all external sub-terms.
+Additional desugarings are shown in the supplement and a technique based on \cite{TSLs} could be introduced to allow tycon providers to define more such desugarings  composably.
 
@@
 \end{figure*}
-
-\subsection{Generalized Introductory Operations}\label{sec:introop}
-The translation of the generalized introductory form, $\eintro{\sttmidx}{\es}$, is determined by the tycon of the type it is being analyzed against as a function of the type's index, the \emph{term index}, $\sttmidx$, and the \emph{argument list}, $\es$.
-
-Before discussing rules (ana-intro) and (ana-intro-other), we note that we can recover a variety of standard concrete introductory forms by a purely syntactic desugaring to this abstract form (and thus allow their use at more than one type).
-For example, for regular strings we will be able to use the string literal form, $\texttt{"s"}$, which desugars to $\eintro{\texttt{"s"}_\text{SL}}{\cdot}$.
-The term index is the corresponding static value of kind $\keyw{Str}$, indicated by a subscript for clarity.
-Similarly, for labeled products, records, objects and so on, we can define a generalized labeled collection form, $\{\mathtt{lbl}_1=e_1, \ldots, \mathtt{lbl}_n=e_n\}$, that desugars to $\eintro{\texttt{[}\mathtt{lbl}_1, \ldots, \mathtt{lbl}_n\texttt{]}}{e_1; \ldots; e_n}$, i.e.
-a list constructed from the row labels is the term index and the corresponding row values are the arguments.
-In both cases, the term index captures  static portions of the concrete form and the arguments capture all external sub-terms.
-Additional desugarings are shown in the supplement and a technique based on \cite{TSLs} could be introduced to allow tycon providers to define more such desugarings  composably.
-
 Let us derive $\small\eana{\Upsilon_\text{ex}}{\Phi_\text{rstr}\Phi_\text{lprod}}{e_\text{ex}}{\stx{paper}}{\iota_\text{ex}}$ where we define an example typing context $\small\Upsilon_\text{ex}  := title \Rightarrow \stx{title}$ and $\small e_\text{ex} := \{\conclbl{title}=title, \conclbl{conf}=\texttt{"EXMPL 2015"}\}$.
@@
 This is a static function that is applied, in the seventh premise, to determine whether the term is well-typed, raising an error if not or determining the translation of the term if so.
-The function has access to the type index, the term index and an interface to the list of arguments.
-The  judgement $\vdash_\Phi \omega \sim \psi$, which appeared as the final premise of the rule (tcc-ext),  checks that the intro opcon definition is well-kinded: rule (ocstruct-intro) in Figure \ref{ocstruct}.
+The function has access to the type index, the term index and an interface to the list of arguments, and its kind is checked by the  judgement $\vdash_\Phi \omega \sim \psi$, which appeared as the final premise of the rule (tcc-ext) and is defined in Figure \ref{ocstruct}.
 
@@
 \quad \keyw{let}~\svar{aok} :: \kunit = \svar{arity0}~\svar{args}~\keyw{in}~\\
-\quad \keyw{let}~\svar{rok} :: \kunit = \svar{rmatch}~\svar{tyidx}~\svar{tmidx}~\keyw{in}~\\
-\quad \svar{str\_of\_Str}~\svar{tmidx}
+\quad \keyw{let}~\svar{rok} :: \kunit = \svar{rmatch}~\svar{tyidx}~\svar{tmidx}~\keyw{in}~\svar{str\_of\_Str}~\svar{tmidx}
 }}}
@@
 
-The only introductory form for kind $\kitm$ is $\sqitm{\qitm}$, where $\qitm$ is a \emph{translational internal term}.
-This form is analagous to the introductory form for kind $\kity$ described previously, $\sqity{\qity}$, where $\qity$ is a translational internal type.
+\paragraph{Term Translations} The only introductory form for kind $\kitm$ is $\sqitm{\qitm}$, where $\qitm$ is a \emph{translational internal term}.
+This form is analagous to the introductory form for kind $\kity$ described in Sec.
+\ref{sec:type-translations}, $\sqity{\qity}$.
 Each form in the syntax of $\iota$ has a corresponding form in the syntax for $\qitm$ and both the kinding rules and operational semantics simply recurse through these in the same manner as shown in Sec.
@@
 These stand in for the translation of argument $n$, the first if it arises via analysis against type $\st$ and the second if it arises via type synthesis.
-They are not intended to be written directly by tycon providers, arising only internally in the semantics of argument interface lists.
+They are not intended to be written directly by tycon providers, arising only internally via argument interface lists.
 Before giving the rules, let us motivate the mechanism with the intro opcon definition for $\tcvar{lprod}$.
@@
 The recursive case first checks that the label provided in the term index matches the label provided in the type index, using a helper function $\svar{lbleq} :: \klbl \rightarrow \klbl \rightarrow \kunit$.
-Then, we request type analysis of the corresponding argument, $\svar{rowarg}$,  against the type in the type index, $\svar{rowty}$, by writing $\svar{ana}~\svar{rowarg}~\svar{rowty}$, which  produces a translational internal term  referring to the translation of that argument if it succeeds, discussed next.
+Then, we request type analysis of the corresponding argument, $\svar{rowarg}$,  against the type in the type index, $\svar{rowty}$, by writing $\svar{ana}~\svar{rowarg}~\svar{rowty}$.
+Here, $\svar{ana}$ is a helper function defined below that produces a translational internal term  of the form $\sqitm{\anatrans{n}{\st}}$ where $\st$ is the value of $\svar{rowty}$ if the row value is well-typed and fails if not.
 The final line constructs a nested tuple based on this translation and the recursive result.
@@
 The argument interface list, called $\stx{args}$ in the rules, is simply a static list of kind $\kargs$ where the $i$th entry is $(\slam{\kty}{\svar{ty}}{\sana{i}{\svar{ty}}}, \slam{\kunit}{\_}{\ssyn{i}})$.
-It is generated by the judgement $\keyw{args}(n)=\stx{args}$, where $n$ is the length of the argument list, written $|\es|=n$ (see supplement).
+It is generated by the judgement $\keyw{args}(n)=\stx{args}$, where $n$ is given as the length of the argument list, written $|\es|=n$ (see supplement).
 
@@
 
-\paragraph{Abstracted Term Translations} The reason that the translations themselves were not inserted yet (even though they were derived in rule (s-ana-success) above) is again because we want to be able to  hold these abstract by replacing them with variables.
+\paragraph{Abstracted Term Translations} The reason for this  is again because we will hold these abstract by replacing them with variables.
 The judgement $\edeabs{\tc}{\argEnv}{\qitm}{\memD}{\memG}{\itm}{\memD'}{\memG'}$, appearing as the eighth premise of (ana-intro), relates a translational internal term $\qitm$ to an internal term $\itm$ called the corresponding \emph{abstracted term translation}, where all references to the translation of an argument (having any type) are replaced with a unique variable and all references to the translation of a type constructed by a user-defined tycon other than the ``delegated tycon'' $\tc$ are replaced with an abstract type variable as described in Sec.
@@
     \eanaX{e}{\st}{\iota}\\
-    \tdeabs{\Phi}{\tc}{\srep{\st}}{\memD}{\ity}{\memD'}\\
+    \tdeabs{\Phi}{c}{\srep{\st}}{\memD}{\ity}{\memD'}\\
     (x~\text{fresh})
 }{
-    \edeabs{\tc}{\es;\Upsilon;\Phi}{\anatrans{n}{\st}}{\memD}{\memG}{x}{\memD'}{\memG, n : \st \leadsto \iota/x : \tau}
+    \edeabs{c}{\es;\Upsilon;\Phi}{\anatrans{n}{\st}}{\memD}{\memG}{x}{\memD'}{\memG, n : \st \leadsto \iota/x : \tau}
 }
@@
 
-The rule (ana-intro-other) is used to introduce values of a type constructed by an ``other'' tycon.
+\paragraph{Other Intro} The rule (ana-intro-other) is used to introduce values of a type constructed by an ``other'' tycon.
 The term index, rather than the tycon context, directly specifies the static function that maps the arguments to a translation.
@@
 All non-introductory opcons associated with user-defined tycons go through another generalized form, in this case for \emph{targeted operations}, $\etarg{\opname{op}}{\sttmidx}{e_\text{targ}}{\es}$, where $\opname{op}$ ranges over opcon names, ${\sttmidx}$ is the term index, $e_\text{targ}$ is the \emph{target argument} and $\es$ are the remaining arguments.
-Some examples of desugarings to this form include explicit invocation, $e_\text{targ}.\opname{op}\langle\sttmidx\rangle(\es)$ (and variants of where the term index or the arguments are omitted), projection syntax for use by record-like types, $e_\text{targ}\opname{\#}\conclbl{lbl}$, which desugars to $\etarg{\opname{\#}}{\conclbl{lbl}}{e_\text{targ}}{\cdot}$, and concatenative syntax $e_\text{targ} \cdot e_\text{arg}$, which desugars to $\etarg{\opname{conc}}{\striv}{e_\text{targ}}{e_\text{arg}}$.
+
+Desugarings to this form include an explicit form, $e_\text{targ}.\opname{op}\langle\sttmidx\rangle(\es)$ (and variants of where the term index or the arguments are omitted), projection syntax for use by record-like types, $e_\text{targ}\opname{\#}\conclbl{lbl}$, which desugars to $\etarg{\opname{prj}}{\conclbl{lbl}}{e_\text{targ}}{\cdot}$, and concatenative syntax $e_\text{targ} \cdot e_\text{arg}$, which desugars to $\etarg{\opname{conc}}{\striv}{e_\text{targ}}{e_\text{arg}}$.
 We show other desugarings, including case analysis, in the supplement.
@@
 The fifth premise of (syn-targ) extracts the targeted opcon definition of $\opname{op}$ from the opcon structure, $\omega$.
-Like the intro opcon definition, this is a static function that generates a translational internal term on the basis of the delegated tycon's type index, the term index and an argument interface list.
-Targeted opcon definitions also synthesize a type.
-The rule (ocstruct-targ) in Fig.
-\ref{ocstruct} ensures that it is well-kinded.
+Like the intro opcon definition, this is a static function that generates a translational internal term on the basis of the target tycon's type index, the term index and an argument interface list.
+Targeted opcon definitions additionally synthesize a type.
+The rule (ocstruct-targ) in Figure \ref{ocstruct} ensures that no tycon defines an opcon twice and that the opcon definitions are well-kinded.
 For example, $\omega_\text{rstr/targops}$ defines:
@@
 \]
-The first line simply checks that exactly two total arguments, including the target term, were passed in using the helper function $\svar{arity2}$.
+The first line of the body checks that exactly two arguments, including the target argument, were passed in via helper function $\svar{arity2}$.
 We then request synthesis of both arguments.
@@
 
-The remaining premises of (syn-targ) are analagous to the corresponding premises in (ana-intro), with the only exception being that we check the abstract term translation against the abstract type translation of the synthesized type.
+The remaining premises of (syn-targ) are analagous to the corresponding premises in (ana-intro), with the only difference being that we check the abstract term translation against the abstract type translation of the synthesized type.
 Like (ana-intro-other), the rule (syn-targ-other) is used when the target synthesizes an ``other'' type.
@@
 \end{theorem}
-
-The case for the form $\ssyn{n}$ requires that the following theorem be simultaneously defined.
-The mutual induction is well-founded because the total length of all the argument lists in the terms being considered monotonically decreases.
+The case in the proof for the form $\ssyn{n}$ requires that the following theorem be simultaneously defined.
+The mutual induction is well-founded because the total length of all the argument lists in the terms being considered decreases.
 \begin{theorem}[Type Synthesis]
@@
 Type safety for the IL \cite{pfpl} then implies that well-typed external terms cannot go wrong.
-To prove this, we must prove a stronger theorem, type-preserving translation (the analog of type-preserving compilation in the typed compilation literature \cite{tarditi+:til-OLD}\todo{right citation for TIL?}).% produced as the translation of the external type.
+To prove this, we must prove a stronger theorem, type-preserving translation (the analog of type-preserving compilation in the typed compilation literature \cite{tarditi+:til-OLD}\todo{right citation for TIL?}):% produced as the translation of the external type.
 Note that we need only to state our top-level theorems for the analysis judgement  because of the subsumption rule.
@@
 We induct on the typing judgement.
-The interesting cases are (ana-intro), (ana-intro-other), (syn-trans) and (syn-trans-other); the later two arise via subsumption.
+The interesting cases are (ana-intro), (ana-intro-other), (syn-trans) and (syn-trans-other); the latter two arise via subsumption.
 The result follows directly from the final premise of each rule, combined with a lemma that states that the substitutions $\gamma$ and $\delta$ generated by the abstracted type and term translation stores satisfy the corresponding $\Delta_\text{abs}$ and $\Gamma_\text{abs}$, and so applying them in the conclusion gives a well-typed term.
@@
 This serves to ensure \emph{hygienic translation} -- translations cannot refer to variables in the surrounding scope directly, so uniformly renaming a variable cannot change the meaning of a program.
-Variables in $\Upsilon$ can (e.g.
-$title$ in the previous example) occur in arguments, but the translations of the arguments only appear \emph{after} the substitution $\gamma$ has been applied.
+Variables in $\Upsilon$ can  occur in arguments (e.g.
+$title$ in the previous example), but the translations of the arguments only appear \emph{after} the substitution $\gamma$ has been applied.
 We assume that substitution is capture-avoiding in the usual manner.
@@
 \cite{sanitation-psp14} and extending this to our  setting requires showing only that the opcon definitions in $\tcvar{rstr}$ adequately capture these specifications using standard program correctness techniques for the SL, which is a simple functional language.
-The rule (syn-targ-other) can synthesize a regular string type, but the abstracted translation will always be checked against $\tau_\text{abs}=\alpha$, so no strings that could not arise by another rule can be generated.
+The only ``wrinkle'' is that the rule (syn-targ-other) can synthesize a regular string type, but the abstracted translation will always be checked against $\tau_\text{abs}=\alpha$, so no strings that could not arise by another rule could  be generated, maintaining the invariant.
 \end{proof-sketch}
@@
 $\stx{paper}$, and generating a translation that is not in the corresponding regular language, e.g.
-$\texttt{""}$ could be defined, but when used, the rule (syn-targ) would check this translation against $\tau_\text{abs}=\alpha$ and the check would fail.
+$\texttt{""}$, could be defined, but when used, the rule (syn-targ) would check this translation against $\tau_\text{abs}=\alpha$ and the check would fail.
 %Type abstraction is the basis for conservatively composing type system fragments, just like it is the basis for composing ML-style modules.
@@
 \end{itemize} then for all $\Phi' = \Phi, \tcdef{\tc'}{\psi'}{\theta'}$ such that $\vdash \Phi'$, the same tycon invariant holds under $\Phi'$: \begin{itemize}
-\item If $\eana{\emptyset}{\Phi'}{e}{\sty{\tc}{\sttyidx}}{\iota}$ and $\vdash_{\Phi'}{\sty{\tc}{\sttyidx}}\leadsto{\tau}$ and $\iota \Downarrow \iota'$ then $P(\sttyidx, \iota)$.
+\item If $\eana{\emptyset}{\Phi'}{e}{\sty{\tc}{\sttyidx}}{\iota}$ and $\vdash_{\Phi'}{\sty{\tc}{\sttyidx}}\leadsto{\tau}$ and $\iota \Downarrow \iota'$ then $P(\sttyidx, \iota')$.
 \end{itemize}
@@
 The proof maps every well-typed term in the extended tycon context to a well-typed term in the original tycon context with the same translation, so we can apply the original invariant to arrive at the property $P(\sttyidx, \iota')$.
-This term is constructed by replacing all types constructed by $\tc'$ with a type constructed by $\keyw{other}[m]$, for an $m$ uniquely associated with $\tc'$, and all introductory and targeted sub-terms associated with $\tc'$ with an equivalent one that passes through the rules (ana-intro-other) and (syn-targ-other).\end{proof-sketch}
+This term is constructed by replacing all types constructed by $\tc'$ with a type constructed by $\keyw{other}[m]$, for an $m$ uniquely associated with $\tc'$, and all introductory and targeted sub-terms associated with $\tc'$ with an equivalent one that passes through the rules (ana-intro-other) and (syn-targ-other) by simply partially applying the intro and targeted opcon definitions to generate the term indices.
+Types constructed by $\keyw{other}[m]$ are indistinguishable from types constructed by a ``future'' tycon because the type case analysis construct is not defined for this tycon form, so the translations coincide.\end{proof-sketch}
 %Had the translational term generated by the intro opcon definition been, e.g., $(\anatrans{0}{\stx{title}}, (\texttt{"TEST"}, ())$, then the corresponding abstract term would be $(x_0, (\texttt{"TEST"}, ())$ and the check would fail because $(\alpha_0, \alpha_1)~(x_0 : \alpha_0) \nvdash \texttt{"TEST"} : \alpha_1$.
@@
 
-\section{Related Work and Discussion}\label{prior-work}
+\section{Related Work}\label{prior-work}
 

### 2014-11-01 att-pldi15/att-pldi15.tex
More revisions through sec5

@@
 $\mathcal{L}\{{\rightharpoonup}\,{\forall}\,{\mu}\,{\keyw{1}}\,{\times}\,{+}\}$ is the language that includes partial function types, polymorphic types, recursive types, nullary and binary product types and binary sum types (its syntax is shown in Figure \ref{syntax-IL}, discussed below).
-Other researchers introduce fragments by defining a simple calculus with a ``catch-all'' constant and base type to stand notionally for all other terms and types that may also be included in some future complete language.\todo{find a really good example, or add this to final PSP rev.??}
+Another related approach is to describe fragments with a simple calculus having a ``catch-all'' constant and base type to stand notionally for all other terms and types that may also be included in some future complete language.
 
@@
 
-This is not an everday problem for programmers only because fragments like those mentioned above are ``general purpose'': they make it possible to \emph{isomorphically embed}  many other fragments as ``libraries''.
-For example, list types need not be built in because they are isomorphic to the type $\iforall{\alpha}{\imu{t}{\iunit + (\alpha \times t))}}$ (e.g.
-datatypes in ML, which combine these into a single declaration construct).
+This is not an everyday problem for programmers only because fragments like those mentioned above are ``general purpose'': they make it possible to \emph{isomorphically embed}  many other fragments as ``libraries''.
+For example, list types need not be built in because they are isomorphic to the type $\iforall{\alpha}{\imu{t}{\iunit + (\alpha \times t))}}$ (datatypes in ML combine these into a single declaration construct).
 %Languages providing datatypes al\'a ML  are perhaps most directly oriented around  embeddings into this core.
@@
 
-Nevertheless, situations do arise where it is not possible to  use these fragments to establish an isomorphic embedding that preserves a desirable fragment's  static and dynamic semantics, including  performance bounds specified by a cost semantics.
+Universality properties (i.e.
+``Turing-completeness'') can be invoked to guarantee that an embedding that preserves a desirable fragment's dynamic semantics is possible, but an isomorphic embedding must also preserve the static semantics and, if defined, performance bounds specified using a cost semantics.
+This is not always possible.
 Embeddings are also sometimes too \emph{complex}, as measured by the cost of the extralinguistic computations that are needed to map in and out of the embedding and, if these  must be performed mentally by programmers, considering various human factors.
@@
 many academic researchers) by making them unavailable to programmers.
-While programmers can choose from dialects supporting, e.g., a principled approach to distributed programming, or one that builds in support for statically reasoning about units of measure, there may not be an available dialect supporting both.
+While programmers can choose from dialects supporting, e.g., a principled approach to distributed programming, or one that builds in support for statically reasoning about units of measure, one that supports both fragments may not be available.
 Using different dialects separately for different components of a program is untenable: components written in different dialects cannot always interface safely (i.e.
@@
 
-These problems do not arise when a fragment is expressed as an isomorphic embedding (i.e.
+These problems do not merit concern when a fragment is expressed as an isomorphic embedding (i.e.
 as a library) because modern \emph{module systems} enforce abstraction barriers that ensure that the isomorphism need only be established in the ``closed world'' of the module.
-There are no ``link-time'' proof obligations for clients in the ``open world''.
-For example, a module defining the semantics of sets in ML can hold the representation of sets abstract, ensuring that any invariants maintained by the functions in the module (e.g.
-uniqueness, if using a list representation) will hold no matter which other modules are separately in use by a client.
-%Mechanisms that can help decrease the complexity of an embedding without violating abstraction barriers are thus valuable, and we will lead into our work in Sec.
+For example, a module defining sets in ML can hold the representation of sets abstract, ensuring that any invariants maintained by the functions in the module (e.g.
+uniqueness, if using a list representation) will hold no matter which other modules are separately in use by a client.\todo{citation?} %Mechanisms that can help decrease the complexity of an embedding without violating abstraction barriers are thus valuable, and we will lead into our work in Sec.
 \ref{prior-work} by summarizing them.
@@
 
-\paragraph{Contributions} In this paper, we take foundational steps towards this goal by  constructing a minimal but powerful core calculus, @$\lambda$ (the ``actively typed'' lambda calculus).
+\paragraph{Contributions} In this paper, we take foundational steps towards this goal by  constructing a simple but surprisingly powerful core calculus, @$\lambda$ (the ``actively typed'' lambda calculus).
 %Despite its minimality, it can host a variety of practical semantic extensions like those described above, while maintaining strong metatheoretic guarantees and, crucially, providing modular reasoning properties.
@@
 \ref{types}.
-We describe how tycons define their associated term-level operator constructors (opcons) in Sec.
+We describe how tycons control the semantics of their associated term-level operator constructors (opcons) in Sec.
 \ref{external-terms}.
@@
 We describe  related work in Sec.
-\ref{prior-work} and conclude in Sec.
+\ref{prior-work} and conclude with a brief discussion of expressiveness in Sec.
 \ref{discussion}.
@@
 Programmers interface with @$\lambda$ by writing \emph{external terms}, $e$.
-The syntax of external terms is shown in Figure \ref{syntax-EL}.
+The abstract syntax of external terms is shown in Figure \ref{syntax-EL} (we will give various concrete desugarings in Sec.
+\ref{external-terms}).
 %We will describe useful syntactic desugarings atop this syntax as we go on and review recent techniques that permit modularly introducing  desugarings like these in Sec.
@@
 Note that our specifications in this paper are intended to be algorithmic: we indicate ``outputs'' when introducing judgement forms by \emph{mode annotations}, $\moutput$; these are not part of the judgement's syntax.
-In particular, note that the type is an ``output'' in the synthetic judgement, but an ``input'' in  the analytic judgement.
+In particular, note that the type is an ``output'' only in the synthetic judgement.
 
@@
 Scala\todo{citation}) because it eliminates the need for type annotations in many situations while remaining decidable in more situations than whole-function type inference and providing what are widely perceived to be higher quality error messages\todo{citation from Dunfield}.
-Secondly, it will give us a clean way to reuse the generalized introductory form, $\eintro{\st}{\es}$, and its associated desugarings, at many types, in a manner that relates to recent mechanisms supporting type-specific syntax extensions \cite{TSLs}.
-For example, when we define regular string types, we will be able to reuse standard string literal syntax.
+Secondly, it will give us a clean way to reuse the generalized introductory form, $\eintro{\st}{\es}$, and its associated desugarings, at many types.
+For example, regular string types will be able to use standard string literal syntax.
 
@@
 
-The internal dynamics are specified also in a standard way as a structural operational semantics with a stepping judgement {\small $\iota \mapsto \iota\moutput$} and a value judgement {$\iota~\mathtt{val}$}.
+The internal dynamics are specified as a structural operational semantics with a stepping judgement {\small $\iota \mapsto \iota\moutput$} and a value judgement {$\iota~\mathtt{val}$}.
 The multi-step judgement $\iota \mapsto^{*} \iota\moutput$ is the reflexive, transitive closure of the stepping judgement and the evaluation judgement $\iota \Downarrow \iota'$ is defined iff $\iota \mapsto^{*} \iota'$ and $\iota'~\mathtt{val}$.
@@
 The reader can consider these as forming a total subset of ML or a simply-typed subset of Coq.
-The semantics also directly follow \cite{pfpl}, so we  omit the details here.
+The semantics also directly follow \cite{pfpl}, so we  omit the details of this core here.
 Only three new  kinds will be needed: $\kty$ (Sec.
@@
 
-The dynamic semantics of static terms is defined as a structural operational semantics by the stepping judgement $\sstep{\st}{\argEnv}{\st\moutput}$, the value judgement $\sval{\st}{\argEnv}$ and the error judgement $\serr{\st}{\argEnv}$.
+The dynamic semantics of static terms is defined as a structural operational semantics by a stepping judgement $\sstep{\st}{\argEnv}{\st\moutput}$, a value judgement $\sval{\st}{\argEnv}$ and an error judgement $\serr{\st}{\argEnv}$.
 Here, $\argEnv$ ranges over \emph{argument environments}, which we will  return to when considering opcons in Sec.
@@
 The introductory form for kind $\kty$ is $\sty{c}{\st}$, where $c$ is a \emph{tycon} and $\st$ is the \emph{type index}.
-The syntax  for tycons in Figure \ref{syntax-TC} specifies that $c$ is either the built-in tycon governing partial functions, $\rightharpoonup$, a user-defined tycon name written in small caps, $\tc$, or an ``other'' tycon, $\keyw{other}[m]$ indexed by a natural number $m$ (to allow arbitrarily many such tycons).
+The syntax  for tycons given in Figure \ref{syntax-TC} specifies that $c$ is either the built-in tycon governing partial functions, $\rightharpoonup$, a user-defined tycon name written in small caps, $\tc$, or an ``other'' tycon, $\keyw{other}[m]$ indexed by a natural number $m$ (to allow arbitrarily many such tycons).
 The kinding rules governing the form $\sty{c}{\st}$ are shown in Figure \ref{fig:types}.
@@
 As our first example, consider the user-defined tycon $\tcvar{rstr}$.
-The index kind of $\tcvar{rstr}$ is $\krx$, which classifies static regular expressions and is defined as an inductive sum kind in the usual way.
+The index kind of $\tcvar{rstr}$ is $\krx$, which classifies static regular expression patterns and we assume is defined as an inductive sum kind in the usual way.
 Types constructed by $\tcvar{rstr}$ will classify \emph{regular strings}, which are statically known to be in the regular language specified by the type index  \cite{sanitation-psp14}.
-For example, $\stx{title} := \sty{\tcvar{rstr}}{\concrx{.+}}$ classifies non-empty strings and $\stx{conf} := \sty{\tcvar{rstr}}{\concrx{[A-Z]+ \digit\digit\digit\digit}}$ classifies conference names.
+For example, $\stx{title} := \sty{\tcvar{rstr}}{\concrx{.+}}$ will classify non-empty strings and $\stx{conf} := \sty{\tcvar{rstr}}{\concrx{[A-Z]+ \digit\digit\digit\digit}}$ will classify conference names.
 The type indices are  here written using standard concrete syntax for concision.
@@
 
-\subsection{Type Case Analysis}
-A type $\st$ can be case analyzed against a known tycon $c$ using $\stycase{c}{\st}{\sx}{\st_1}{\st_2}$.
+\subsection{Static Type Case Analysis}
+A type $\st$ can be case analyzed using $\stycase{c}{\st}{\sx}{\st_1}{\st_2}$.
 If the value of $\st$ is constructed by $c$, its type index is bound to $\sx$ and the branch $\st_1$ is taken.
@@
 % \end{mathpar}
-
-The rule for arrow types is analagous, but no rule for tycons of the form $\keyw{other}[m]$ is defined.
+The rule for $c{=}{\rightharpoonup}$ is analagous, but no rule for $c{=}\keyw{other}[m]$ is defined (these always take the default branch).
 
@@
 \label{fig:tycon-ctxs}
+\vspace{-5px}
 \end{figure}
@@
 
-\subsection{Type Equivalence} 
+\subsection{Type Equivalence}\label{sec:type-equivalence}
 The first check simplifies the handling of type equivalence: type index kinds must be \emph{equality kinds}, i.e.
@@
 \subsection{Type Translations}\label{sec:type-translations}
-Each tycon {computes} translations for the types it constructs as a function of each type's index by specifying a \emph{translation schema} as the first component of the {tycon structure}, $\theta$.
-For a tycon with index kind $\ktyidx$, the translation schema must have kind $\ktyidx \rightarrow \kity$, checked by (tcc-ext).
+Each tycon in the tycon context {computes} translations for the types it constructs as a function of each type's index by specifying a \emph{translation schema} in the {tycon structure}, $\theta$.
+A tycon with index kind $\ktyidx$ defines a translation schema of kind $\ktyidx \rightarrow \kity$, checked by (tcc-ext).
 
@@
 
-The kinding rules and operational semantics for these shared forms simply proceed recursively, e.g.
+The kinding rules (and dynamics) for these shared forms simply proceed recursively, e.g.
 \begin{mathpar}
@@
 
-The syntax for translational internal types additionally includes an ``unquote'' form,  $\qtuq{\st}$, so that they can be constructed compositionally, as well as a form, $\srep{\st}$, that allows one type's translation to refer to the translation of another type $\st$.
-\begin{mathpar}\small
+The syntax for translational internal types additionally includes an ``unquote'' form,  $\qtuq{\st}$, so that they can be constructed compositionally, as well as a form, $\srep{\st}$, that allows one type's translation to refer to the translation of another type $\st$: \begin{mathpar}\small
 \inferrule[k-ity-unquote]{
@@
 
-We choose a translation schema for $\tcvar{lprod}$ that generates nested binary product types by folding over the type index and referring to the translations of the types therein (though this is not the only workable choice, e.g.
+We choose a translation schema for $\tcvar{lprod}$ that generates nested binary product types by recursing over the type index and referring to the translations of the types therein (though this is not the only workable choice, e.g.
 we could also have used a list).
-We assume $\small\svar{listfold} :: \kforall{\kalpha_1}{\kforall{\kalpha_2}{\klist{\kalpha_1}\rightarrow\kalpha_2\rightarrow(\kalpha_1\rightarrow\kalpha_2\rightarrow\kalpha_2)\rightarrow\kalpha_2}}$ in defining $\theta_\text{lprod} := \keyw{trans}=\stx{lprod/trans}, \omega_\text{lprod}$  where $\stx{lprod/trans}:=$
+We assume $\small\svar{listrec} :: \kforall{\kalpha_1}{\kforall{\kalpha_2}{\klist{\kalpha_1}\rightarrow\kalpha_2\rightarrow(\kalpha_1\rightarrow\kalpha_2\rightarrow\kalpha_2)\rightarrow\kalpha_2}}$ in defining $\theta_\text{lprod} := \keyw{trans}=\stx{lprod/trans}, \omega_\text{lprod}$  where $\stx{lprod/trans}:=$
 \[\small
 \begin{array}{l}
-\slam{\klist{\klbl \times \kty}}{\svar{tyidx}}{\skap{\kity}{\skap{\klbl\times\kty}{\svar{listfold}}}~\svar{tyidx}~\sqity{\iunit}~\\
+\slam{\klist{\klbl \times \kty}}{\svar{tyidx}}{\skap{\kity}{\skap{\klbl\times\kty}{\svar{listrec}}}~\svar{tyidx}~\sqity{\iunit}~\\
   \quad (\lambda \svar{h}{:}\klbl \times \kty.\lambda \svar{r}{:}\kity.\sqity{\srep{{\ssnd{\svar{h}}}}\times\qtuq{\svar{r}}}}
@@
 $\memD_\text{paper/abs}  := \stx{title} \leftrightarrow \keyw{str}/\alpha_1, \stx{conf} \leftrightarrow \keyw{str}/\alpha_2$.
-The type translation store $\memD ::= \emptyset ~|~ \memD, \st \leftrightarrow \ity/\alpha$ maintains the  mapping from types, $\st$, to their (non-abstract) translations, $\tau$, and the type variables, $\alpha$, which appear in their place.
+The \emph{type translation store} $\memD ::= \emptyset ~|~ \memD, \st \leftrightarrow \ity/\alpha$ maintains the  mapping from types, $\st$, to their (non-abstract) translations, $\tau$, and the type variables, $\alpha$, which appear in place of their translation.
 
@@
 \ref{overview}.
-We simply determine any selectively abstract translation, then apply the substitution:
+We simply determine any selectively abstract translation, then apply the induced substitution:
 \begin{mathpar}\small
@@
 With this intuition, we can now give the rules for the selective type abstraction judgement.
-We recurse generically over sub-terms of $\qity$ until sub-terms of form $\srep{\st}$ are encountered (see supplement).
-
-The translation of partial function types is direct and is not held abstract, so that lambdas can be used as the sole binding construct in the external language:
+It first recurses generically over sub-terms of $\qity$ until sub-terms of form $\srep{\st}$ are encountered (see supplement).
+The translation of partial function types is direct and is not held abstract, so that lambdas can be used as the sole binding construct in the EL:
 \begin{mathpar}
@@
 \end{mathpar}
-The translation of a user-defined types constructed by the delegated tycon is determined by calling the translation schema and checking that the type translation it generates refers only to type variables generated from $\memD'$:
+The translation of a user-defined type constructed by the delegated tycon is determined by calling the translation schema and checking that the type translation it generates refers only to type variables generated from $\memD'$:
 \begin{mathpar}
@@
 \inferrule[abs-tc-local]{
-    \tcdef{\tc}{\psi}{\tcstruct{\stx{schema}}{\omega}} \in \Phi\\\\
-    \sevalNA{\stx{schema}(\sttyidx)}{\sqity{\qity}}\\
-    \tdeabs{\Phi}{\tc}{\qity}{\memD}{\ity}{\memD'}\\\\
+    \tcdef{\tc}{\psi}{\tcstruct{\stx{schema}}{\omega}} \in \Phi\\
+    \sevalNA{\stx{schema}~\sttyidx}{\sqity{\qity}}\\
+    \tdeabs{\Phi}{\tc}{\qity}{\memD}{\ity}{\memD'}\\
     \memD' \leadsto \delta : \Delta\\
@@
 }{
-    \tdeabs{\Phi}{c}{\srep{\sty{\tc}{\sttyidx}}}{\memD}{\alpha}{\memD', \sty{\tc}{\sttyidx} \leftrightarrow \tau/\alpha}
-}
-\end{mathpar}
-The translation of an ``other'' type is given directly in its index (the rule where it is held abstract is in the supplement):
+    \tdeabs{\Phi}{c}{\srep{\sty{\tc}{\sttyidx}}}{\memD}{\alpha}{\memD', \sty{\tc}{\sttyidx} \leftrightarrow [\delta]\tau/\alpha}
+}
+\end{mathpar}
+The translation of an ``other'' type is given directly in its index (rule (abs-other-foreign-new) is in the supplement):
 \begin{mathpar}
 \small
-\inferrule[abs-tc-other-new]{
+\inferrule[abs-other-local]{
     \tdeabs{\Phi}{\keyw{other}[m]}{\qity}{\memD}{\tau}{\memD'}
@@
 Because we are defining a bidirectional type system, a subsumption rule is needed to allow synthetic terms to be analyzed against an equivalent type.
-Per above, equivalent types must be  syntactically identical at normal form, and we consider analysis only if $\istype{\st}{\Phi}$, so the rule (subsume) is straightforward.
+Per Sec.
+\ref{sec:type-equivalence}, equivalent types must be  syntactically identical at normal form, and we consider analysis only if $\istype{\st}{\Phi}$, so the rule (subsume) is straightforward.
 To use an analytic term in a synthetic position, the programmer must provide a type ascription, written $e : \st$.
@@
 
-Variables and functions behave in the standard manner given our definitions of types and type translations (used to generate the ascription on the lambda in the IL, which is intrinsically rather than bidirectionally typed).
+Variables and functions behave in the standard manner given our definitions of types and type translations (used to generate ascriptions in the IL, which is intrinsically rather than bidirectionally typed).
 We use Plotkin's fixpoint operator for general recursion (cf.
-\cite{pfpl}), and define both only in analytic positions for simplicity.
+\cite{pfpl}), and define both lambdas and fixpoints only in analytic positions for simplicity.
 
@@
 Before discussing rules (ana-intro) and (ana-intro-other), we note that we can recover a variety of standard concrete introductory forms by a purely syntactic desugaring to this abstract form (and thus allow their use at more than one type).
-For example, for regular strings we will be able to use the string literal form, $\texttt{"s"}$, which desugars to $\eintro{\texttt{"s"}_\text{SL}}{\cdot}$.
+For example, for regular strings we can use the string literal form, $\texttt{"s"}$, which desugars to $\eintro{\texttt{"s"}_\text{SL}}{\cdot}$.
 The term index is the corresponding static value of kind $\keyw{Str}$, indicated by a subscript for clarity.
@@
 \caption{Checking opcon structures against tycon signatures.}
-\label{ocstruct}
+\label{ocstruct}\vspace{-5px}
 \end{figure*}
-Let us derive $\small\eana{\Upsilon_\text{ex}}{\Phi_\text{rstr}\Phi_\text{lprod}}{e_\text{ex}}{\stx{paper}}{\iota_\text{ex}}$ where we define an example typing context $\small\Upsilon_\text{ex}  := title \Rightarrow \stx{title}$ and $\small e_\text{ex} := \{\conclbl{title}=title, \conclbl{conf}=\texttt{"EXMPL 2015"}\}$.
+Let us derive $\small\eana{\Upsilon_\text{ex}}{\Phi_\text{rstr}\Phi_\text{lprod}}{e_\text{ex}}{\stx{paper}}{\iota_\text{ex}}$ where $\small\Upsilon_\text{ex}  := title \Rightarrow \stx{title}$ and $\small e_\text{ex} := \{\conclbl{title}=title, \conclbl{conf}=\texttt{"EXMPL 2015"}\}$.
 The  translation will be $\small\iota_\text{ex} := (title, (\texttt{"EXMPL 2015"}_\text{IL}, ()))$, where $\small\texttt{"EXMPL 2015"}_\text{IL}$ is an internal string (of internal type $\keyw{str}$).
@@
 
-The second premise extracts the \emph{intro term index kind}, $\klitidx$, from the \emph{opcon signature}, $\chi$, of the tycon signature and the third premise checks the provided term index against this kind.
+The second premise extracts the \emph{intro term index kind}, $\klitidx$, from the \emph{opcon signature}, $\chi$, and the third premise checks the provided term index against this kind.
 This is simply the kind of term index expected by the tycon, e.g., $\tcvar{lprod}$ specifies $\klist{\klbl}$, so that it can use the labeled collection form, while $\tcvar{rstr}$ specifies an intro index kind of $\keyw{Str}$, so that it can use the string literal form.
@@
 It begins by making sure that no arguments were passed in (we will return to arguments and the fifth and sixth premises of (ana-intro) with the next example), using the helper function $\svar{arity0} :: \kargs \rightarrow \kunit$ defined such that any non-empty list will raise an error, via the static term $\sraise{\kunit}$.
-In practice, the tycon provider would specify an error message here.%The kinding rules and operational semantics of this exception-like form are standard:
-
+In practice, the tycon provider would specify an error message here.
 Next, it checks the string provided as the term index against the regular expression given as the type index using $\svar{rmatch} :: \krx \rightarrow \keyw{Str} \rightarrow \kunit$, which we assume is defined in the usual way and again raises an error on failure.
@@
 
-\paragraph{Term Translations} The only introductory form for kind $\kitm$ is $\sqitm{\qitm}$, where $\qitm$ is a \emph{translational internal term}.
+The only introductory form for kind $\kitm$ is $\sqitm{\qitm}$, where $\qitm$ is a \emph{translational internal term}.
 This form is analagous to the introductory form for kind $\kity$ described in Sec.
 \ref{sec:type-translations}, $\sqity{\qity}$.
-Each form in the syntax of $\iota$ has a corresponding form in the syntax for $\qitm$ and both the kinding rules and operational semantics simply recurse through these in the same manner as shown in Sec.
+Each form in the syntax of $\iota$ has a corresponding form in the syntax for $\qitm$ and both the kinding rules and dynamics simply recurse through these in the same manner as shown in Sec.
 \ref{sec:type-translations}.
-Also analagously, there is an unquote form, $\quq{\st}$, that permits translational internal terms to be constructed compositionally.
-The supplement gives the analagous rules.
+There is also an analagous unquote form, $\quq{\st}$.
+%The supplement gives the analagous rules.
 
@@
 These stand in for the translation of argument $n$, the first if it arises via analysis against type $\st$ and the second if it arises via type synthesis.
-They are not intended to be written directly by tycon providers, arising only internally via argument interface lists.
 Before giving the rules, let us motivate the mechanism with the intro opcon definition for $\tcvar{lprod}$.
-We have $\omega_\text{lprod} := \tcstructn{\stx{lprod/intro}}; \omega_\text{lprod/targops}$ where $\stx{lprod/intro} :=$
+We have $\omega_\text{lprod} := \tcstructn{\stx{lprod/intro}}; \omega_\text{lprod/targops}$ where $\stx{lprod/intro} :=$\vspace{-5px}
 \[\small
@@
 Then, we request type analysis of the corresponding argument, $\svar{rowarg}$,  against the type in the type index, $\svar{rowty}$, by writing $\svar{ana}~\svar{rowarg}~\svar{rowty}$.
-Here, $\svar{ana}$ is a helper function defined below that produces a translational internal term  of the form $\sqitm{\anatrans{n}{\st}}$ where $\st$ is the value of $\svar{rowty}$ if the row value is well-typed and fails if not.
+Here, $\svar{ana}$ is a helper function defined below that produces a translational internal term  of the form $\sqitm{\anatrans{n}{\st}}$, where $\st$ is the value of $\svar{rowty}$, if the row value is well-typed, and raises an error if not.
 The final line constructs a nested tuple based on this translation and the recursive result.
@@
 
- Recall that the kinding judgement is indexed by $n$, which is an upper bound on the argument index of terms of the form $\sana{n}{\st}$ and $\ssyn{n}$. This is enforced in the kinding rules:
+ Recall that the kinding judgement is indexed by $n$, which is an upper bound on the argument index of terms of the form $\sana{n}{\st}$ and $\ssyn{n}$. This is enforced in their kinding rules:
 \begin{mathpar}\small
@@
 This is to prevent out-of-bounds errors -- tycon providers do not write these forms directly, but only access them via the argument interface list, guaranteed to have the correct length.
-This is possible because:
+This is  safe because:
 \begin{lemma}
@@
 
-The kinding rules also prevent these translational internal term forms generated by these static operations from being well-kinded when $n = 0$.
+The kinding rules also prevent the translational internal term forms generated by these static operations from being well-kinded when $n = 0$.
 Like the translational internal type form $\srep{\st}$, these  are retained in values of kind $\kitm$:
@@
 \inferrule[s-itm-anatrans-v]{
-    \svalNA{\st}\\
+    \sval{\st}{\es;\Upsilon;\Phi}\\
     \keyw{nth}[n](\es)=e
@@
 \paragraph{Abstracted Term Translations} The reason for this  is again because we will hold these abstract by replacing them with variables.
-The judgement $\edeabs{\tc}{\argEnv}{\qitm}{\memD}{\memG}{\itm}{\memD'}{\memG'}$, appearing as the eighth premise of (ana-intro), relates a translational internal term $\qitm$ to an internal term $\itm$ called the corresponding \emph{abstracted term translation}, where all references to the translation of an argument (having any type) are replaced with a unique variable and all references to the translation of a type constructed by a user-defined tycon other than the ``delegated tycon'' $\tc$ are replaced with an abstract type variable as described in Sec.
+The judgement $\edeabs{\tc}{\argEnv}{\qitm}{\memD}{\memG}{\itm\moutput}{\memD'\moutput}{\memG'\moutput}$, appearing as the eighth premise of (ana-intro), relates a translational internal term $\qitm$ to an internal term $\itm$ called the corresponding \emph{abstracted term translation}, where all references to the translation of an argument (having any type) are replaced with a unique variable and all references to the translation of a type constructed by a user-defined tycon other than the ``delegated tycon'' $\tc$ are replaced with an abstract type variable as described in Sec.
 \ref{sec:selective-type-translation-abstraction}.
@@
 The rules for the abstracted term translation judgement follow recursively for shared forms just like those for the abstracted type translation judgement (see supplement).
-The rules for references to argument translations derived via type analysis operates as follows ($\syntrans{n}$ is analagous):
+The rule for references to argument translations derived via type analysis operates as follows ($\syntrans{n}$ is analagous):
 \begin{mathpar}\small
@@
 
-The tenth premise of (ana-intro) determines an abstract type translation for the type provided for analysis, in this case $\small\tau_\text{ex/abs} := \alpha_0 \times (\alpha_1 \times \kunit)$ (alpha-equivalent to $\tau_\text{conf/abs}$ in Sec.
+The tenth premise of (ana-intro) determines an abstract type translation for the type provided for analysis, in this case $\small\tau_\text{ex/abs} := \alpha_0 \times (\alpha_1 \times \kunit)$ (alpha-equivalent to $\tau_\text{paper/abs}$ in Sec.
 \ref{sec:selective-type-translation-abstraction}), and the eleventh premise extracts a type substitution, $\delta_\text{ex/abs}$, and type formation context, $\Delta_\text{ex/abs}$, from $\memD_\text{ex/abs}$, again equivalent to $\delta_\text{conf/abs}$ and $\Delta_\text{conf/abs}$ from Sec.
@@
 
-\paragraph{Other Intro} The rule (ana-intro-other) is used to introduce values of a type constructed by an ``other'' tycon.
+\paragraph{Other Intro} The rule (ana-intro-other) is used to introduce terms of a type constructed by an ``other'' tycon.
 The term index, rather than the tycon context, directly specifies the static function that maps the arguments to a translation.
@@
 
-Desugarings to this form include an explicit form, $e_\text{targ}.\opname{op}\langle\sttmidx\rangle(\es)$ (and variants of where the term index or the arguments are omitted), projection syntax for use by record-like types, $e_\text{targ}\opname{\#}\conclbl{lbl}$, which desugars to $\etarg{\opname{prj}}{\conclbl{lbl}}{e_\text{targ}}{\cdot}$, and concatenative syntax $e_\text{targ} \cdot e_\text{arg}$, which desugars to $\etarg{\opname{conc}}{\striv}{e_\text{targ}}{e_\text{arg}}$.
+Desugarings include an explicit form, $e_\text{targ}.\opname{op}\langle\sttmidx\rangle(\es)$ (and variants of where the term index or the arguments are omitted), projection syntax for use by record-like types, $e_\text{targ}\opname{\#}\conclbl{lbl}$, which desugars to $\etarg{\opname{prj}}{\conclbl{lbl}}{e_\text{targ}}{\cdot}$, and concatenative syntax $e_\text{targ} \cdot e_\text{arg}$, which desugars to $\etarg{\opname{conc}}{\striv}{e_\text{targ}}{e_\text{arg}}$.
 We show other desugarings, including case analysis, in the supplement.
@@
 Whereas introductory operations were analytic, targeted operations are synthetic in @$\lambda$.
-The type and translation is determined by the tycon of the type synthesized by the target argument.
+The type and translation are determined by the tycon of the type synthesized by the target argument.
 The rule (syn-targ) is otherwise similar to (ana-intro) in its structure.
@@
 The opcons associated with $\tcvar{lprod}$ are also straightforward: $\opname{prj}$ projects out the row with the provided label, $\opname{conc}$ concatenates two labeled products (updating common fields with the value of the right argument), and $\opname{drop}$ creates a new labeled product from the target with some fields dropped.
-Note that both regular strings and labeled products define concatenation.
-Targeted operations use no new mechanisms, so we will only show regular string concatenation here; others are in the supplement.
+Note that both regular strings and labeled products can define concatenation without conflict.
+%Targeted operations use no new mechanisms, so we only show regular string concatenation; others are in the supplement.
 
@@
 \]
-The first line of the body checks that exactly two arguments, including the target argument, were passed in via helper function $\svar{arity2}$.
+The first line checks that exactly two arguments, including the target argument, were passed in via helper function $\svar{arity2}$.
 We then request synthesis of both arguments.
@@
 The case in the proof for the form $\ssyn{n}$ requires that the following theorem be simultaneously defined.
-The mutual induction is well-founded because the total length of all the argument lists in the terms being considered decreases.
+The mutual induction is well-founded because the total number of argument lists in the terms being considered decreases.
 \begin{theorem}[Type Synthesis]
@@
 Type safety for the IL \cite{pfpl} then implies that well-typed external terms cannot go wrong.
-To prove this, we must prove a stronger theorem, type-preserving translation (the analog of type-preserving compilation in the typed compilation literature \cite{tarditi+:til-OLD}\todo{right citation for TIL?}):% produced as the translation of the external type.
+To prove this, we must prove a stronger theorem, \emph{type-preserving translation}, which is the analog of type-preserving compilation in the typed compilation literature \cite{tarditi+:til-OLD}\todo{right citation for TIL?}:% produced as the translation of the external type.
 Note that we need only to state our top-level theorems for the analysis judgement  because of the subsumption rule.
@@
 The interesting cases are (ana-intro), (ana-intro-other), (syn-trans) and (syn-trans-other); the latter two arise via subsumption.
-The result follows directly from the final premise of each rule, combined with a lemma that states that the substitutions $\gamma$ and $\delta$ generated by the abstracted type and term translation stores satisfy the corresponding $\Delta_\text{abs}$ and $\Gamma_\text{abs}$, and so applying them in the conclusion gives a well-typed term.
+The result follows directly from the final premise of each rule, combined with lemmas that states that if $\memD \leadsto \delta : \Delta_\text{abs}$ and $\memG \leadsto \gamma : \Gamma_\text{abs}$ then $\vdash \delta : \Delta$ and $\Delta~\Gamma \vdash \gamma : \Gamma_\text{abs}$, i.e.
+that all the variables in the contexts have well-formed/well-typed substitutions, and thus that applying them in the conclusions of the rules gives a well-typed term.
 \end{proof-sketch}
@@
 \paragraph{Hygienic Translation} 
-Note in the just mentioned rules that the domains of $\Upsilon$ and $\Gamma_\text{abs}$ are disjoint.
+Note above that the domains of $\Upsilon$ (and thus $\Gamma$)  and $\Gamma_\text{abs}$ are disjoint.
 This serves to ensure \emph{hygienic translation} -- translations cannot refer to variables in the surrounding scope directly, so uniformly renaming a variable cannot change the meaning of a program.
@@
 \begin{theorem}[Unicity]
-If $\vdash \Phi$ and $\vdash_\Phi \Upsilon \leadsto \Gamma$ and $\vdash_\Phi \st \leadsto \tau$ and $\vdash_\Phi \st' \leadsto \tau'$ and $\eanaX{e}{\st}{\iota}$ and $\eanaX{e}{\st'}{\iota'}$ then $\st = \st'$ and $\iota = \iota'$.
+If $\vdash \Phi$ and $\vdash_\Phi \Upsilon \leadsto \Gamma$ and $\vdash_\Phi \st \leadsto \tau$ and $\vdash_\Phi \st' \leadsto \tau'$ and $\eanaX{e}{\st}{\iota}$ and $\eanaX{e}{\st'}{\iota'}$ then $\st = \st'$ and $\tau=\tau'$ and $\iota = \iota'$.
 \end{theorem}
@@
 \cite{sanitation-psp14} and extending this to our  setting requires showing only that the opcon definitions in $\tcvar{rstr}$ adequately capture these specifications using standard program correctness techniques for the SL, which is a simple functional language.
-The only ``wrinkle'' is that the rule (syn-targ-other) can synthesize a regular string type, but the abstracted translation will always be checked against $\tau_\text{abs}=\alpha$, so no strings that could not arise by another rule could  be generated, maintaining the invariant.
+The only ``wrinkle'' is that the rule (syn-targ-other) can synthesize a regular string type.
+Luckily the abstracted translation will always be checked against $\tau_\text{abs}=\alpha$, so no strings that could not arise by another rule could  be generated, by the standard abstraction theorem for the IL.
 \end{proof-sketch}
 
-The reason why the rule (syn-targ-other) is not a problem in proving a tycon invariant turns out to be the same reason extending the tycon context also cannot violate such a tycon invariant.
-Any newly introduced tycon defining a targeted operator that synthesizes a regular string type, e.g.
+The reason why the rule (syn-targ-other) is never a problem in proving a tycon invariant turns out to be the same reason extending the tycon context conserves all tycon invariants.
+A newly introduced tycon defining a targeted operator that synthesizes a regular string type, e.g.
 $\stx{paper}$, and generating a translation that is not in the corresponding regular language, e.g.
-$\texttt{""}$, could be defined, but when used, the rule (syn-targ) would check this translation against $\tau_\text{abs}=\alpha$ and the check would fail.
+$\texttt{""}$, could be defined, but when used, the rule (syn-targ) would check the translation against $\tau_\text{abs}=\alpha$, so the invariant could not be broken by the abstraction theorem for the IL.
 %Type abstraction is the basis for conservatively composing type system fragments, just like it is the basis for composing ML-style modules.
- We can state this conservativity property more generally.
+ We can state this conservativity property more generally:
 

### 2014-11-02 att-pldi15/att-pldi15-supplement.tex
Finished draft of PLDI15 except some citations

--- 2014-11-01 att-pldi15/att-pldi15.tex
@@
 \title{Modularly Composing Typed Language Fragments}
-
+\subtitle{Supplemental Material}
 \authorinfo{}{}{}
@@
 \maketitle
-\begin{abstract}
-Researchers often describe typed programming languages as fragments or simple calculi, leaving to language designers the task of composing these to form complete languages.
-%comparable to those available for library-based embeddings packaged using a modern module system.
-This is not a systematic process: metatheoretic results must be established anew for each composition, guided only notionally by metatheorems derived for simpler systems.
-As the design space grows, mechanisms that provide stronger modular reasoning principles than this are needed.%, so that important metatheoretic results need only be established locally and, ultimately, so that language extensions can be treated like libraries.
-%,  Sometimes this fails because certain language features do not ``play well'' together, but precisely characterizing which is elusive given these practices.
-
-%Recent work has started to address this problem, e.g.
-by showing how to modularly reason about desugarings atop a fixed type system.
-Our focus here is on safely composing type system fragments.
-We organize fragments around type constructors, as is usual practice when describing type systems, and sidestep the difficulties of composing abstract syntax (i.e.
-the expression problem)  by delegating control over a small, uniform abstract syntax in a type-directed manner to logic associated with a type constructor.
-
-In this paper, we begin from first principles with a core calculus, @$\lambda$, specified like many full-scale languages: as a bidirectionally typed translation semantics.
-Only the $\rightarrow$ type constructor (tycon) is built in; all other external tycons (we show a variant of record types and constrained string types) are defined by extending a \emph{tycon context}.
-Each tycon defines the semantics of associated term-level operators (e.g.
-record projection) using functions written in a static language where types and translations are values.
-The semantics provide  strong metatheoretic guarantees, notably \emph{type safety} and \emph{conservativity}: that all \emph{tycon  invariants} will be conserved under extension.
-Mechanized proofs are not needed: problems are caught during typechecking by lifting typed compilation techniques into the semantics and enforcing abstraction barriers around tycons using type abstraction, the same principle that underlies ML-style modules.
-\end{abstract}
 
@@
 %extensible languages; module systems; type abstraction; typed compilation; type-level computation
-\section{Introduction}\label{intro}
-Typed programming languages are most often described as being composed of \emph{fragments}, each contributing to the language's concrete syntax, abstract syntax, static semantics and dynamic semantics.
-In his textbook, Harper organizes fragments around type constructors, describing each in a different chapter \cite{pfpl}.
-Languages are then identified by a set of type constructors, e.g.
-$\mathcal{L}\{{\rightharpoonup}\,{\forall}\,{\mu}\,{\keyw{1}}\,{\times}\,{+}\}$ is the language that includes partial function types, polymorphic types, recursive types, nullary and binary product types and binary sum types (its syntax is shown in Figure \ref{syntax-IL}, discussed below).
-Another related approach is to describe fragments with a simple calculus having a ``catch-all'' constant and base type to stand notionally for all other terms and types that may also be included in some future complete language.
-
-In contrast, the usual metatheoretic reasoning techniques for programming languages  (e.g., rule induction) operate on complete language specifications.
-Each {combination} of fragments must formally be treated as its own monolithic language for which  metatheorems must be established anew, guided only informally by those derived for the smaller systems from which the language is notionally composed.
-
-%In both cases, it is left as an extrinsic concern to ensure that the metatheory developed separately is actually conserved when fragments defined in these ways are composed to form a complete language.
-
-This is not an everyday problem for programmers only because fragments like those mentioned above are ``general purpose'': they make it possible to \emph{isomorphically embed}  many other fragments as ``libraries''.
-For example, list types need not be built in because they are isomorphic to the type $\iforall{\alpha}{\imu{t}{\iunit + (\alpha \times t))}}$ (datatypes in ML combine these into a single declaration construct).
-%Languages providing datatypes al\'a ML  are perhaps most directly oriented around  embeddings into this core.
-% Leaving the core language simple makes it easier to establish its metatheory and verify that tools, like compilers, are implemented correctly.% They also have encouraging connections to logic.
-%For example, we do not need to define the type constructor $\fvar{list}$ (indexed by a type) as a fragment (though it is possible to do so) because there is a user-defined datatype constructor, $\texttt{list}$, parameterized by a type, that is isomorphic.
-If we added the $\fvar{list}$ fragment to the language, it would be entirely redundant semantically: every well-typed term of a type constructed by $\fvar{list}$ corresponds to a well-typed term and a polymorphic recursive sum type such th%Although strings and numbers can be embedded as recursive sum types, it is recognized that this is impractical, so these are also usually included as primitives.
-%Establishing an isomorphic embedding of a desirable fragment in terms of general-purpose fragments is not always possible, nor are such embeddings always practical.
-
-Universality properties (i.e.
-``Turing-completeness'') can be invoked to guarantee that an embedding that preserves a desirable fragment's dynamic semantics is possible, but an isomorphic embedding must also preserve the static semantics and, if defined, performance bounds specified using a cost semantics.
-This is not always possible.
-Embeddings are also sometimes too \emph{complex}, as measured by the cost of the extralinguistic computations that are needed to map in and out of the embedding and, if these  must be performed mentally by programmers, considering various human factors.
-%We will discuss specific examples below.
-%When an embedding is too complex, abstraction providers have a few options, discussed in Sec. \ref{prior-work}. When an embedding is not possible, however, or when these options are  insufficient, 
-%providers are often compelled to introduce these fragments by extending an existing language, thereby forming a new \emph{dialect}.
-To save effort, they may do so by forking existing artifacts or leverage tools like compiler generators or language frameworks, also reviewed in Sec.
-\ref{prior-work}.
-%Indeed, the proliferation of language dialects constructed for this reason might be taken as an evidence that the core language is not as ``general'', when considered comprehensively, as might be hoped.
-Each time a language is extended with a new fragment for one of these reasons, a new \emph{dialect} is born. Within the ML lineage, for example, dialects abound:
-%Reynolds, in a remark that recalls the ``Turing tarpit'' of Perlis \cite{Perl82a}, summarizes the issue \cite{Reynolds94anintroduction}: 
-%\begin{quote}
-%To say that any reasonable function can be expressed by some program is not to say that it can be expressed by the most reasonable program.
-It is clear that the language requires a novel programming style.
-Moreover, it is likely that certain important functions cannot be expressed by their most efficient algorithms.
-%\end{quote}
-%
-
-\begin{enumerate}
-\vspace{-5px}
-%\compresslist
-\item 
-\textbf{General Purpose Fragments:} 
-A number of variations on product types, for example, have been introduced in dialects: 
-$n$-ary tuples, 
-labeled tuples, 
-records (identified up to reordering), 
-records with width and depth subtyping \cite{Cardelli:1984:SMI:1096.1098}, 
-records with update operators\footnote{The Haskell wiki notes that ''No, extensible records are not implemented in GHC.
-The problem is that the record design space is large, and seems to lack local optima.
-[...] As a result, nothing much happens.'' \cite{GHCFAQ}} \cite{ocaml-manual},
-records with mutable fields \cite{ocaml-manual}, 
-%field delegation \cite{atlang-gpce14} \todo{gpce submission} 
-and 
-records with ``methods'' (i.e. pure objects \cite{TSLs}). 
- Sum-like types are also exposed in various ways: 
-finite datatypes, 
-open datatypes \cite{conf/ppdp/LohH06}, 
-hierarchically open datatypes \cite{journals/toplas/MillsteinBC04}, 
-polymorphic variants \cite{ocaml-manual} and 
-ML-style exception types. Combinations of these manifest themselves as class-based object systems \cite{ocaml-manual}.
-
-\item
-\textbf{Specialized Fragments:} Fragments that track specialized static invariants to provide stronger correctness or security guarantees, manage unwieldy lower-level abstractions and run-time systems or control cost are also frequently introduced in dialects, e.g.
-for data parallelism  \cite{chakravarty2007data}, distributed programming \cite{Murphy:2007:TDP:1793574.1793585}, reactive programming \cite{mandel2005reactiveml}, authenticated data structures \cite{Miller:2014:ADS:2535838.2535851}, databases \cite{Ohori:2011:MSM:2034773.2034815},  units of measure \cite{conf/cefp/Kennedy09} and regular string sanitation \cite{sanitation-psp14}.
-%\verb|sprintf| in the OCaml dialect statically distinguishes format strings from strings.% All of these are implemented as dialects of existing languages, presumably because a strong encoding was not feasible.
-
-\item
-\textbf{Foreign Fragments:} A safe and natural foreign function interface (FFI) can be valuable (particularly given this proliferation of dialects).
-This requires enforcing the type system of the foreign language in the calling language.
-%Using  a FFI that does not do this can lead to safety issues, even when both languages are separately known to be safe.
-%Safe FFIs generally require direct extensions to the language. 
-For example, MLj builds in a safe FFI to Java \cite{Benton:1999:IWW:317636.317791}.
-\end{enumerate}
-\vspace{-5px}
-
-This \emph{dialect-oriented} state of affairs is unsatisfying.
-%Language designers are burdened with needing to understand the complete metatheory of every fragment they add to their language to make sure it is not incompatible with existing fragments, so languages change rarely.
-This decreases the impact of many potentially useful innovations designed by fragment providers (e.g.
-many academic researchers) by making them unavailable to programmers.
-While programmers can choose from dialects supporting, e.g., a principled approach to distributed programming, or one that builds in support for statically reasoning about units of measure, one that supports both fragments may not be available.
-Using different dialects separately for different components of a program is untenable: components written in different dialects cannot always interface safely (i.e.
-a safe FFI, item 3 above, is needed).
-
-These problems do not merit concern when a fragment is expressed as an isomorphic embedding (i.e.
-as a library) because modern \emph{module systems} enforce abstraction barriers that ensure that the isomorphism need only be established in the ``closed world'' of the module.
-For example, a module defining sets in ML can hold the representation of sets abstract, ensuring that any invariants maintained by the functions in the module (e.g.
-uniqueness, if using a list representation) will hold no matter which other modules are separately in use by a client.\todo{citation?} %Mechanisms that can help decrease the complexity of an embedding without violating abstraction barriers are thus valuable, and we will lead into our work in Sec.
-\ref{prior-work} by summarizing them.
-
-When library-based embeddings are not possible, as in the examples above, mechanisms are needed that make it possible to define and  reason in a similarly modular manner about  direct extensions to the semantics of a language.
-Such a mechanism could ultimately be integrated directly into the language, blurring the distinction between fragments and libraries and decreasing the need for new dialects.% Importing fragments that introduce new semantics   would be as easy and reasonable as importing a new module is in ML today.
-%In the limit, the community could rely on modularly mechanized metatheory and compiler correctness results.% rather than requiring heroic efforts from individual research groups that consider an entire language at once.
-%Recent work has shown progress on modularly introducing new concrete syntax, reviewed in Sec.
-\ref{desugaring}.
-Our focus is on the problem of introducing new semantics.
-
-\paragraph{Contributions} In this paper, we take foundational steps towards this goal by  constructing a simple but surprisingly powerful core calculus, @$\lambda$ (the ``actively typed'' lambda calculus).
-%Despite its minimality, it can host a variety of practical semantic extensions like those described above, while maintaining strong metatheoretic guarantees and, crucially, providing modular reasoning properties.
-%a calculus introduced briefly in recent work on \emph{active type constructors}  \cite{atlang-gpce14}\todo{TR?
-Arxiv?}, reviewed in Sec.
-\ref{overview}.
-Its semantics are structured like those of many modern languages, consisting of an \emph{external language} (EL) governed by a {typed translation semantics} targeting a much simpler \emph{internal language} (IL).
-Rather than building in a monolithic set of external type constructors, however, the  semantics are indexed by a \emph{tycon context}.
-Each tycon defines the semantics of its operators via functions written in a \emph{static language} (SL) where types and translations are values.
-%, which provides a form of type-level computation.
-%The authors demonstrate the expressive power of this technique primarily with an implementation and a number of examples of fragments as libraries, and their core calculus .% (unlike systems that treat the type system as a ``bag of rules'', where non-determinism can arise).
-
-We will begin by giving an overview of the organization  and main judgements of @$\lambda$ in Sec.
-\ref{atlam}, then discuss how types are constructed and introduce our two main examples, one defining labeled product types with a functional update operator, and the other regular string types, based on a recent core calculus style specification \cite{sanitation-psp14}, in Sec.
-\ref{types}.
-We describe how tycons control the semantics of their associated term-level operator constructors (opcons) in Sec.
-\ref{external-terms}.
-We next give the key metatheoretic properties of the calculus in Sec.
-\ref{metatheory}, including \emph{type safety} and a key modularity result, which we call  \emph{conservativity}: any invariants that can be established about all values of a type under \emph{some} tycon context (i.e.
-in some  ``closed world'') are conserved in any further extended tycon context (i.e.
-in the ``open world'').
-Interestingly, type system providers need not necessarily provide mechanized proofs to maintain these guarantees.
-Instead, the approach we take relies on type abstraction in the internal language.
-As a result, we are able to borrow the same results that underly modular reasoning in simply-typed languages like ML to reason modularly about typed language fragments.
-We describe  related work in Sec.
-\ref{prior-work} and conclude with a brief discussion of expressiveness in Sec.
-\ref{discussion}.
-
-\section{Overview of @$\lambda$}\label{atlam}\label{overview}
-\paragraph{External Language}
-Programmers interface with @$\lambda$ by writing \emph{external terms}, $e$.
-The abstract syntax of external terms is shown in Figure \ref{syntax-EL} (we will give various concrete desugarings in Sec.
-\ref{external-terms}).
-%We will describe useful syntactic desugarings atop this syntax as we go on and review recent techniques that permit modularly introducing  desugarings like these in Sec.
-\ref{desugaring}.
-Our main focus here is on semantic extensions.
-The static and dynamic semantics are given simultaneously as a \emph{bidirectionally typed translation semantics}, i.e.
-the key judgements take the form:  \[\esynX{e}{\st\moutput}{\iota\moutput} ~~~~~\text{and}~~~~~ \eanaX{e}{\st}{\iota\moutput}\]
-\noindent
-
-These are pronounced ``$e$ (synthesizes / analyzes against) type $\sigma$ and has  translation $\iota$ under typing context $\Upsilon$ and tycon context $\Phi$''.
-Note that our specifications in this paper are intended to be algorithmic: we indicate ``outputs'' when introducing judgement forms by \emph{mode annotations}, $\moutput$; these are not part of the judgement's syntax.
-In particular, note that the type is an ``output'' only in the synthetic judgement.
-
-This basic separation of the EL and IL is commonly used for full-scale language specifications, e.g.
-the Harper-Stone semantics for Standard ML \cite{Harper00atype-theoretic}.
-The internal language is purposely kept small, e.g.
-defining only simple products, to simplify metatheoretic reasoning and compilation.
-The EL then specifies various useful higher-level constructs, e.g.
-record types, by translation to the IL.
-In @$\lambda$, the EL builds in only function types.
-All other external constructs are defined in the \emph{tycon context}, described in the Sec.
-\ref{types}.
-
-We choose bidirectional typechecking, also sometimes called \emph{local type inference} \cite{Pierce:2000:LTI:345099.345100}, for two main reasons.
-The first is once again to justify the practicality of our approach: local type inference is increasingly being used in modern languages (e.g.
-Scala\todo{citation}) because it eliminates the need for type annotations in many situations while remaining decidable in more situations than whole-function type inference and providing what are widely perceived to be higher quality error messages\todo{citation from Dunfield}.
-Secondly, it will give us a clean way to reuse the generalized introductory form, $\eintro{\st}{\es}$, and its associated desugarings, at many types.
-For example, regular string types will be able to use standard string literal syntax.
-
-Unlike the Harper-Stone semantics, where external and internal terms were governed by a common type system, in @$\lambda$ each external type, $\st$, maps onto an internal type, $\tau$, called the \emph{type translation} of $\st$.
-This mapping is specified by the  type translation judgement, $\vdash_\Phi \st \leadsto \tau$, which will be described in Sec.
-\ref{sec:type-translations}.
-For this reason, this specification style may also be compared to specifications for the first stage of a type-directed compiler, e.g.
-the TIL compiler for Standard ML \cite{tarditi+:til-OLD}, here lifted ``one level up'' into the semantics of the language itself.
-As we will see, type safety follows from a property analagous to a correctness condition that arises in typed compilers.
-Modular reasoning will be based on holding the type translation of $\st$ abstract ``outside'' the tycon.
-
-%We will return to how we control the semantics of the single introductory form in the abstract syntax, $\eintro{\st}{\es}$, on the basis of the type it is being analyzed against, sidestepping issues related to the \emph{expression problem} as a result \todo{cite other paper}\cite{wadler1998expression}.
-The form $\eother{\iota}$ is a technical device used to quantify over all terms that may arise in a ``future'' tycon context, serving an analagous role to the  ``catchall'' base constants commonly found in minimal calculi, and we will return to its semantics later as well.%We will see how we leverage this bidirectionality when we discuss literal forms below.
-% The judgement is only well-defined for \emph{valid} tycon contexts, written judgementally as $\vdash \Phi$ and \emph{valid} typing contexts, $\vdash_\Phi \Upsilon$, which we also specify below.
-%  \begin{figure*}[t!]
-% \small
-% \hspace{-5px}\input{example-table}
-% %\\~\\
-% %$\begin{array}{rcl}
-% %e\cdot\conclbl{lbl}(\conclbl{lbl}_1=e_1, \cdots, \conclbl{lbl}_n=e_n)  :=  \etarg{(\conclbl{lbl}; \svar{list}n[\klbl]~\conclbl{lbl}_1~\cdots~\conclbl{lbl}_n)}{e}{e_1; \cdots; e_n}
-% %\end{array}$
-% \caption{An example external term, $e_{ex}$, in concrete syntax (left), desugared to abstract syntax (right), with static terms shown in green in examples (only).
-The notation ${\small \scolor{\desugar{...}}}$ stands for an embedding of the indicated ${\small \scolor{\conclbl{label}}}$, ${\small \scolor{\concrx{regular expression}}}$, ${\small \scolor{\concstr{string}}}$ or numeral into the SL, with derived kinds  $\klbl$, $\krx$,  $\kstr$ and $\knat$, not shown.
-The signatures of helper functions, e.g.
-${\small \scolor{\svar{nil}}}$ and ${\small \scolor{\svar{list}n}}$, are shown in Figure \ref{helper-sigs}.}
-% \label{example}
-% \end{figure*}
-\begin{figure}[t]
-\small
-\hspace{-5px}\input{syntax-table-IL}
-\caption{Syntax of {$\mathcal{L}\{{\rightharpoonup}\,{\forall}\,{\mu}\,{\iunit}\,{\times}\,{+}\}$}, our internal language (IL).
-Metavariable $x$ ranges over term variables and $\alpha$ and $t$ both range over type variables.}
-\label{syntax-IL}
-\vspace{-10px}
-\end{figure}
-
-\begin{figure}[t]
-\small
-\hspace{-5px}\input{syntax-table-EL}
-\caption{Syntax of the external language (EL).}\label{syntax-EL}
-\end{figure}
-
-\paragraph{Internal Language} 
-%Internal terms, $\iota$, together with internal types, $\tau$, form the \emph{typed internal language}. 
-@$\lambda$ relies on a typed internal language supporting type abstraction (i.e.
-universal quantification over types).
-We use {$\mathcal{L}\{{\rightharpoonup}\,{\forall}\,{\mu}\,{\iunit}\,{\times}\,{+}\}$}, the syntax for which is shown in Figure \ref{syntax-IL}, as representative of any intermediate language for a typed functional language.
-%Note that the IL has a fixed semantics; changes to the IL induce different dialects of @$\lambda$.
-
-We assume the statics of the IL are specified in the standard way by judgements for  type formation {$\iDelta \vdash \tau$}, typing context formation { $\iDelta \vdash \iGamma$} and type assignment {$\iDelta~\iGamma \vdash \iota : \tau\moutput$}.
-%The typing and type formation contexts obey standard structural properties (i.e. weakening, exchange and contraction). 
-In examples, we will omit leading $\emptyset$, used as the base case for finite mappings, and $\cdot$, used as the base case for finite sequences (e.g. writing $\Gamma_\text{test} := x : \tau$).
-
-The internal dynamics are specified as a structural operational semantics with a stepping judgement {\small $\iota \mapsto \iota\moutput$} and a value judgement {$\iota~\mathtt{val}$}.
-The multi-step judgement $\iota \mapsto^{*} \iota\moutput$ is the reflexive, transitive closure of the stepping judgement and the evaluation judgement $\iota \Downarrow \iota'$ is defined iff $\iota \mapsto^{*} \iota'$ and $\iota'~\mathtt{val}$.
-Both the static and dynamic semantics of the IL can be found in any standard textbook covering typed lambda calculi (we directly follow \cite{pfpl}), so we assume familiarity and key metatheoretic properties.
-
-% We also here define a syntax for simultaneous substitutions, of terms for variables, $\gamma$, and of types for polymorphic type variables, $\delta$, and assume standard judgements ensuring that these are valid with respect to a valid context, $\Delta \vdash \gamma : \Gamma$ and $\vdash \delta : \Delta$.
-We apply these substitutions to terms, types and typing contexts using the syntax $[\gamma]\iota$, $[\delta]\iota$, $[\delta]\tau$ and $[\delta]\Gamma$ (in some prior work, application of a substitution like this is written using a hatted form, e.g.
-$\hat\gamma(\iota)$; we intend the same semantics but use a notation more consistent with standard substitution, e.g.
-$[\iota/x]\iota'$).
-We will omit  leading $\emptyset$ and $\cdot$ in examples; in specifications, the former is used for  metatheoretic finite mappings, the latter for metatheoretic ordered lists.
-
-\paragraph{Static Language}
-\begin{figure}[t]
-\small
-\hspace{-5px}\input{syntax-table-SL}
-\caption{Syntax of the static language (SL). Metavariable $\sx$ ranges over static term variables, $\kalpha$ and $\kvar$ over kind variables and $m$ and $n$ over natural numbers.}\vspace{-5px}
-\label{syntax-SL}
-\end{figure}
-The workhorse of @$\lambda$ is the \emph{static language}, which itself forms a typed lambda calculus where 
-%External terms are classified by (external) \emph{types}. 
-\emph{kinds}, $\kappa$, classify \emph{static terms}, $\sigma$.
-The syntax of the SL is given in Figure \ref{syntax-SL}.
-The portion of the SL covered by the first row of kinds and first four rows of static terms in the syntax forms an entirely standard  functional programming language consisting of total functions, universal quantification over kinds, inductive kinds (constrained by the standard positivity condition to prevent non-termination), and products and sums.
-The reader can consider these as forming a total subset of ML or a simply-typed subset of Coq.
-The semantics also directly follow \cite{pfpl}, so we  omit the details of this core here.
-Only three new  kinds will be needed: $\kty$ (Sec.
-\ref{types}), $\kity$ (Sec.
-\ref{sec:type-translations}) and $\kitm$ (Sec.
-\ref{sec:introop}).
-
-The kinding judgement takes the form $\sofkX{\st}{\kappa\moutput}$, where $\kDelta$ and $\kGamma$ are analagous to $\iDelta$ and $\iGamma$ and analagous kind and kinding context formation judgements $\kDelta \vdash \kappa$ and $\kDelta \vdash \kGamma$ are  defined.
-The natural number $n$ is used as a technical device in our semantics to ensure that the forms shown as being indexed by $n$ in the syntax only arise in a controlled manner internally to prevent ``out of bounds'' issues, as we will discuss; they would have no corresponding concrete syntax so $n$ can be assumed $0$ in user-defined terms.
-
-The dynamic semantics of static terms is defined as a structural operational semantics by a stepping judgement $\sstep{\st}{\argEnv}{\st\moutput}$, a value judgement $\sval{\st}{\argEnv}$ and an error judgement $\serr{\st}{\argEnv}$.
-Here, $\argEnv$ ranges over \emph{argument environments}, which we will  return to when considering opcons in Sec.
-\ref{external-terms}.
-The multi-step judgement $\smanystep{\st}{\argEnv}{\st\moutput}$ is the reflexive, transitive closure of the stepping judgement.
-The normalization judgement $\seval{\st}{\argEnv}{\st'}$ is defined iff $\smanystep{\st}{\argEnv}{\st'}$ and $\sval{\st'}{\argEnv}$.
-%${\st}{\memD}{\memG}{\aCtx}{\st\moutput}{\memD\moutput}{\memG\moutput}$, where $\memD$, $\memG$ and $\aCtx$ are also technical devices that will be described later; they too can be ignored from the perspective of user code.
-We write $\st \Downarrow \st'$ iff $\snorm{\st}{\emptyset}{\cdot}{{\rightharpoonup}; \emptyset; \emptyset}{\st'}{\emptyset}{\cdot}$.
-\emph{Static values} are normalized static terms.
-Normalization can also raise an error (to indicate a type error in an external term, or a problem in a tycon definition, as we will discuss), indicated by the judgement $\serrX{\st}$.
-We omit error propagation rules.%We will refer to the relevant rules as we proceed.
-\begin{figure}[t]
-\small
-$\begin{array}{lrcl}
-\textbf{tycons} & c & ::= & \rightharpoonup ~|~ \tc ~|~ \keyw{other}[m]\\
-\textbf{tycon contexts} & \Phi & ::= & \cdot ~|~ \Phi, \tcdef{\tc}{\psi}{\theta}\\
-\textbf{tycon structures} & \theta & ::= & \tcstruct{\st}{\omega} \\
-\textbf{tycon sigs} & \psi & ::= & \tcsig{\kappa}{\chi}\\
-\textbf{opcon structures} & \omega & ::= & \tcstructn{\st} ~|~ \tcstructc{\omega}{op}{\st}\\
-\textbf{opcon sigs}& \chi & ::= & \introsig{\kappa} ~|~ \opsigS{\theta}{op}{\kappa}\\
-\end{array}$
-\caption{Syntax of tycons. Metavariables $\tc$ and $\opname{op}$ range over user-defined tycon and opcon names, respectively, and $m$ ranges over natural numbers.}
-\label{syntax-TC}
-\end{figure}
-\begin{figure*}\begin{mathpar}
-\small
-\inferrule[k-parr]{
-    \sofkX{\st}{\kprod{\kty}{\kty}}
-}{
-    \sofkX{\sty{\rightharpoonup}{\st}}{\kty}
-}
-
-\inferrule[k-ty]{
-    \tcdef{\tc}{\tcsig{\ktyidx}{\chi}}{\theta} \in \Phi\\
-    \sofkX{\st}{\ktyidx}
-}{
-    \sofkX{\sty{\tc}{\st}}{\kty}
-}
-
-\inferrule[k-otherty]{
-    \sofkX{\sttyidx}{\keyw{Nat} \times \kity}
-}{
-    \sofkX{\sty{\keyw{other}[m]}{\sttyidx}}{\kty}
-}\vspace{-10px}
-\end{mathpar}
-\caption{Kinding rules for types, which take the form $\sty{c}{\sttyidx}$ where $c$ is a tycon and $\sttyidx$ is the type index.}
-\label{fig:types}
-\end{figure*}
-
-\section{Types}\label{types}
-
-External types, or simply \emph{types}, are static values of kind $\kty$.
-The introductory form for kind $\kty$ is $\sty{c}{\st}$, where $c$ is a \emph{tycon} and $\st$ is the \emph{type index}.
-The syntax  for tycons given in Figure \ref{syntax-TC} specifies that $c$ is either the built-in tycon governing partial functions, $\rightharpoonup$, a user-defined tycon name written in small caps, $\tc$, or an ``other'' tycon, $\keyw{other}[m]$ indexed by a natural number $m$ (to allow arbitrarily many such tycons).
-The kinding rules governing the form $\sty{c}{\st}$ are shown in Figure \ref{fig:types}.
-The dynamics are simple (see supplement): the index is eagerly normalized and errors propagate.
-We write $\istype{\st}{\Phi}$  iff $\sofkn{\emptyset}{\emptyset}{\Phi}{0}{\st}{\kty}$ and $\svalNA{\st}$.
-
-The rule \rulename{k-parr} specifies that the type index of partial function types must be a pair of types.
-We thus say that $\rightharpoonup$ has \emph{index kind} $\kty \times \kty$.
-To recover a conventional syntax, we can introduce a desugaring from $\st_1 \rightharpoonup \st_2$ to $\sty{\rightharpoonup}{(\st_1, \st_2)}$.
-%Here, we will benefit by treating it uniformly.
-
-All user-defined tycons must be defined in the \emph{tycon context}, $\Phi$, which is simply a list of tycon definitions.
-Each tycon defines a name, $\tc$, a \emph{tycon structure}, $\theta$, and a \emph{tycon signature}, $\psi$.
-We will return to the tycon structure below.
-Tycon signatures have the form $\tcsig{\ktyidx}{\chi}$, where $\ktyidx$ is the tycon's index kind and $\chi$ is the \emph{opcon signature}, which we   discuss in Sec.
-\ref{external-terms}.
-The first premise of \rulename{k-ty} extracts the index kind and the second checks the type index against it.
-
-The rule (k-otherty) governs types constructed by an ``other'' tycon.
-These will serve only as technical devices to stand in for types other than those in a given tycon context in Sec.
-\ref{metatheory}.
-The index of such a type must pair a natural number with a type translation of kind $\kity$, discussed in Sec.
-\ref{sec:type-translations}.
-
-\paragraph{Examples}
-As our first example, consider the user-defined tycon $\tcvar{rstr}$.
-The index kind of $\tcvar{rstr}$ is $\krx$, which classifies static regular expression patterns and we assume is defined as an inductive sum kind in the usual way.
-Types constructed by $\tcvar{rstr}$ will classify \emph{regular strings}, which are statically known to be in the regular language specified by the type index  \cite{sanitation-psp14}.
-For example, $\stx{title} := \sty{\tcvar{rstr}}{\concrx{.+}}$ will classify non-empty strings and $\stx{conf} := \sty{\tcvar{rstr}}{\concrx{[A-Z]+ \digit\digit\digit\digit}}$ will classify conference names.
-The type indices are  here written using standard concrete syntax for concision.
-Previous work has shown how to define type-specific (here, kind-specific) syntax like this composably in libraries \cite{TSLs}.
-We define a tycon context containing only the definition of $\tcvar{rstr}$, $\Phi_\text{rstr} := \tcdef{\tcvar{rstr}}{\tcsig{\krx}{\chi_\text{rstr}}}{\theta_\text{rstr}}$.
-
-Our second example is the tycon $\tcvar{lprod}$, which will define a variant of labeled product type (labeled products are like record types, but maintain a row  ordering; record types are also definable in a manner discussed in the supplement, but maintaining an ordering simplifies our discussion).
-We choose the index kind of $\tcvar{lprod}$ to be $\klist{\klbl \times \kty}$, where list kinds are defined as inductive sums in the usual way, and $\klbl$ classifies static representations of row labels.
-The tycon context containing only $\tcvar{lprod}$'s definition is $\small\Phi_\text{lprod} := \tcdef{\tcvar{lprod}}{\tcsig{\klist{\klbl \times \kty}}{\chi_\text{lprod}}}{\theta_\text{lprod}}$.
-
-In the tycon context containing both tycon definitions, $\Phi_\text{rstr}\Phi_\text{lprod}$, we can define a labeled product type classifying conference papers, $\stx{paper} := \sty{\tcvar{lprod}}{\{\conclbl{title} : \stx{title}, \conclbl{conf} : \stx{conf}\}}$.
-Note that $\istype{\stx{paper}}{\Phi_\text{rstr}\Phi_\text{lprod}}$ and we again use kind-specific syntax, in this case for $\klbl$ and $\klist{\klbl \times \kty}$.
-
-\subsection{Static Type Case Analysis}
-A type $\st$ can be case analyzed using $\stycase{c}{\st}{\sx}{\st_1}{\st_2}$.
-If the value of $\st$ is constructed by $c$, its type index is bound to $\sx$ and the branch $\st_1$ is taken.
-For totality, a default branch, $\st_2$, must also be provided.
-For example, the kinding rule for when $c$ is user-defined is below.
-\begin{mathpar}
-\small
-\inferrule[k-tycase]{
-    \sofkX{\st}{\kty}\\
-    \tcdef{\tc}{\tcsig{\ktyidx}{\chi}}{\theta} \in \Phi\\
-    \sofk{\kDelta}{\kGamma, \sx :: \ktyidx}{\Phi}{\st_1}{\kappa}\\
-    \sofkX{\st_2}{\kappa}
-}{
-    \sofkX{\stycase{\tc}{\st}{\sx}{\st_1}{\st_2}}{\kappa}
-}
-\end{mathpar}
-% \begin{mathpar}\small
-% \inferrule[s-tycase-fail-2]{ }{
-%     \sstep{\stycase{c}{\sotherty{m}{\tau}}{x}{\st_1}{\st_2}}{\argEnv}{\st_2}
-% }
-% \end{mathpar}
-The rule for $c{=}{\rightharpoonup}$ is analagous, but no rule for $c{=}\keyw{other}[m]$ is defined (these always take the default branch).
-
-Put another way, types can be thought of as arising from a distinguished ``open datatype'' defined by the tycon context\todo{citation}.
-%The dynamics (see supplement) are again straightforward, consistent with this intuition.% We will see an example of its use in Sec.
-\ref{sec:targops}.
-\begin{figure}\hfill \fbox{$\vdash \Phi$}\vspace{-25px}\begin{mathpar}
-\small
-\inferrule[tcc-ext]{
-    \vdash \Phi\\
-    \tc \notin \text{dom}(\Phi)\\
-    \keq{\emptyset}{\ktyidx}\\
-    \sofkz{\emptyset}{\emptyset}{\Phi}{\stx{schema}}{\ktyidx \rightarrow \kity}\\\\
-    \vdash_{\Phi,  \tcdef{\tc}{\tcsig{\ktyidx}{\chi}}{\tcstruct{\stx{schema}}{\omega}}} \omega \sim \tcsig{\ktyidx}{\chi}
-}{
-    \vdash \Phi, \tcdef{\tc}{\tcsig{\ktyidx}{\chi}}{\tcstruct{\stx{schema}}{\omega}}
-}\vspace{-8px}
-\end{mathpar}
-\caption{Tycon context well-definedness. We omit the trivial case for when $\Phi=\cdot$ for concision.}
-\label{fig:tycon-ctxs}
-\vspace{-5px}
-\end{figure}
+
+\section{Internal Language}
+\todo{copy from one of the 312 HWs}
+
+\subsection{Substitutions}
+
+\subsection{Abstraction Theorem}
+\todo{hm...}
+
+\section{Tycon Contexts}
 \subsection{Tycon Context Well-Definedness}
-The tycon context well-definedness judgement, $\vdash \Phi$, shown in Figure \ref{fig:tycon-ctxs}, requires that all tycon names are unique and performs additional checks,  described below.
-
-\subsection{Type Equivalence}\label{sec:type-equivalence}
-The first check simplifies the handling of type equivalence: type index kinds must be \emph{equality kinds}, i.e.
-those for which semantic equivalence implies syntactic equality at normal form.
-We define these by the judgement $\keq{\kDelta}{\kappa}$ (see supplement).
-Equality kinds are similar to equality types as found in Standard ML\todo{citation}. The main implication of this choice is that type indices cannot contain static functions.
-
-\subsection{Type Translations}\label{sec:type-translations}
-Each tycon in the tycon context {computes} translations for the types it constructs as a function of each type's index by specifying a \emph{translation schema} in the {tycon structure}, $\theta$.
-A tycon with index kind $\ktyidx$ defines a translation schema of kind $\ktyidx \rightarrow \kity$, checked by (tcc-ext).
-
-The kind $\kity$ has a single introductory form, $\sqity{\qity}$, where $\qity$ is a \emph{translational internal type}.
-Each form in the syntax for internal types, $\tau$,  corresponds to a form in the syntax of translational internal types, $\qity$.
-For example, our translation schema for $\tcvar{rstr}$  simply chooses to ignore the type index and represent all regular strings internally as strings, of internal type abbreviated $\keyw{str}$.
-We abbreviate the corresponding translational internal type $\hat{\keyw{str}}$ and define the translation schema as $\small\theta_\text{rstr} := \tcstruct{\slam{\krx}{\svar{tyidx}}{\sqity{\hat{\keyw{str}}}}}{\omega_\text{rstr}}$.
-
-The kinding rules (and dynamics) for these shared forms simply proceed recursively, e.g.
-\begin{mathpar}
-\small
-\inferrule[k-ity-prod]{
-    \sofkX{\sqity{\qity_1}}{\kity}\\
-    \sofkX{\sqity{\qity_2}}{\kity}
-}{
-    \sofkX{\sqity{\qity_1 \times \qity_2}}{\kity}
-}
-\end{mathpar}
-%The operational semantics for shared forms are also straightforwardly recursive (see supplemental material).
-
-The syntax for translational internal types additionally includes an ``unquote'' form,  $\qtuq{\st}$, so that they can be constructed compositionally, as well as a form, $\srep{\st}$, that allows one type's translation to refer to the translation of another type $\st$: \begin{mathpar}\small
-\inferrule[k-ity-unquote]{
-    \sofkX{\st}{\kity}   
-}{
-    \sofkX{\sqity{\qtuq{\st}}}{\kity}
-}
-
-\inferrule[k-ity-trans]{
-    \sofkX{\st}{\kty}
-}{
-    \sofkX{\sqity{\srep{\st}}}{\kity}
-}
-\end{mathpar}
-The unquote form is eliminated during normalization, while references to the translation of a type are retained in values of kind $\kity$. The key rules in the dynamics are: 
-\begin{mathpar}
-\small
-\inferrule[s-ity-unquote-elim]{
-    \sval{\sqity{\qity}}{\argEnv}
-}{
-    \sstep{\sqity{\qtuq{\sqity{\qity}}}}{\argEnv}{\sqity{\qity}}
-}
-
-\inferrule[s-ity-trans-val]{
-    \sval{\st}{\argEnv}
-}{
-    \sval{\sqity{\srep{\st}}}{\argEnv}
-}
-\end{mathpar}
-
-We choose a translation schema for $\tcvar{lprod}$ that generates nested binary product types by recursing over the type index and referring to the translations of the types therein (though this is not the only workable choice, e.g.
-we could also have used a list).
-We assume $\small\svar{listrec} :: \kforall{\kalpha_1}{\kforall{\kalpha_2}{\klist{\kalpha_1}\rightarrow\kalpha_2\rightarrow(\kalpha_1\rightarrow\kalpha_2\rightarrow\kalpha_2)\rightarrow\kalpha_2}}$ in defining $\theta_\text{lprod} := \keyw{trans}=\stx{lprod/trans}, \omega_\text{lprod}$  where $\stx{lprod/trans}:=$
-\[\small
-\begin{array}{l}
-\slam{\klist{\klbl \times \kty}}{\svar{tyidx}}{\skap{\kity}{\skap{\klbl\times\kty}{\svar{listrec}}}~\svar{tyidx}~\sqity{\iunit}~\\
-  \quad (\lambda \svar{h}{:}\klbl \times \kty.\lambda \svar{r}{:}\kity.\sqity{\srep{{\ssnd{\svar{h}}}}\times\qtuq{\svar{r}}}}
-\end{array}\]
-Evaluating this translation schema with the index of $\stx{paper}$, for example, produces the value $\stx{paper/trans} := \sqity{\qity_\text{paper/trans}}$ where $\small\qity_\text{paper/trans} := \srep{\stx{title}}\times(\srep{\stx{conf}}\times \iunit)$.
-Note that we do not include logic to optimize away the trailing unit type for simplicity (and to again emphasize that many translations are possible for any given type).%For simplicity, and to make the point that the choice of representation has only tycon-local implications, our translation schema does not optimize away
-
-\subsubsection{Selective Type Translation Abstraction}\label{sec:selective-type-translation-abstraction}
-References to translations of other types are maintained in values of kind $\kity$ like this to  allow us to selectively hold them abstract, which will be the key to our main results.
-This can be thought of as analagous to the process in ML by which the true identity of an abstract type in a module is held abstract outside the module until after typechecking.
-The judgement $\small\tdeabs{\Phi}{c}{\qity}{\memD}{\ity\moutput}{\memD\moutput}$ relates a normalized translational internal type $\qity$ to an internal type $\tau$, called a corresponding \emph{abstracted type translation} because references to translations of types constructed by a tycon other than those constructed by a ``delegated tycon'', $c$, are replaced by universally quantified type variables, $\alpha$.
-For example, $\small\tdeabs{\Phi_\text{rstr}\Phi_\text{lprod}}{\tcvar{lprod}}{\qity_\text{paper/trans}}{\emptyset}{\tau_\text{paper/abs}}{\memD_\text{paper/abs}}$
-where $\small\tau_\text{paper/abs} := \alpha_1 \times (\alpha_2 \times \iunit)$ and 
-$\memD_\text{paper/abs}  := \stx{title} \leftrightarrow \keyw{str}/\alpha_1, \stx{conf} \leftrightarrow \keyw{str}/\alpha_2$.
-The \emph{type translation store} $\memD ::= \emptyset ~|~ \memD, \st \leftrightarrow \ity/\alpha$ maintains the  mapping from types, $\st$, to their (non-abstract) translations, $\tau$, and the type variables, $\alpha$, which appear in place of their translation.
-
-Each type translation store induces a \emph{type substitution}, $\delta$, and a corresponding internal type formation context, $\Delta$, according to the judgement $\memD \leadsto \delta : \Delta$.
-Type substitutions simply define an $n$-ary substitution for type variables, $\delta ::= \emptyset ~|~ \delta, \tau/\alpha$.
-For example, $\memD_\text{paper/abs} \leadsto \delta_\text{paper/abs} : \Delta_\text{paper/abs}$ where $\delta_\text{paper/abs} := \keyw{str}/\alpha_1, \keyw{str}/\alpha_2$ and $\Delta_\text{paper/abs} := \alpha_1, \alpha_2$.
-We can apply type substitutions to internal types, terms and typing contexts, written $[\delta]\ity$, $[\delta]\iota$ and $[\delta]\Gamma$, respectively.
-For example, $[\delta_\text{paper/abs}]\tau_\text{paper/abs}$ is $\tau_\text{paper} := \keyw{str} \times (\keyw{str} \times \iunit)$, i.e.
-the final type translation of $\stx{paper}$.
-Indeed, we can now give the rule defining the type translation judgement, $\vdash_\Phi \st \leadsto \tau$, mentioned in Sec.
-\ref{overview}.
-We simply determine any selectively abstract translation, then apply the induced substitution:
-\begin{mathpar}\small
-\inferrule[conc-ty-trans]{
-    \istype{\st}{\Phi}\\
-    \tdeabs{\Phi}{\tc}{\srep{\st}}{\emptyset}{\tau}{\memD}\\
-    \memD \leadsto \delta : \Delta
-}{
-    \vdash_\Phi \st \leadsto [\delta]\tau
-}
-\end{mathpar}
-
-%This process of selectively holding the translations of all types constructed by any tycon other than the ``delegated tycon'' abstract will be critical for conservativity to hold below.
-Here, the translation of the type is being held abstract, rather than the type itself, and type translations are computed from indices, rather than declared ``literally''.
-
-With this intuition, we can now give the rules for the selective type abstraction judgement.
-It first recurses generically over sub-terms of $\qity$ until sub-terms of form $\srep{\st}$ are encountered (see supplement).
-The translation of partial function types is direct and is not held abstract, so that lambdas can be used as the sole binding construct in the EL:
-\begin{mathpar}
-\small
-\inferrule[abs-parr]{
-    \tdeabs{\Phi}{c}{\srep{\st_1}}{\memD}{\ity_1}{\memD'}\\
-    \tdeabs{\Phi}{c}{\srep{\st_2}}{\memD'}{\ity_2}{\memD''}
-}{
-    \tdeabs{\Phi}{c}{\srep{\sty{\rightharpoonup}{(\st_1, \st_2)}}}{\memD}{\tau_1 \rightharpoonup \tau_2}{\memD''}
-}
-\end{mathpar}
-The translation of a user-defined type constructed by the delegated tycon is determined by calling the translation schema and checking that the type translation it generates refers only to type variables generated from $\memD'$:
-\begin{mathpar}
-\small
-\inferrule[abs-tc-local]{
-    \tcdef{\tc}{\psi}{\tcstruct{\stx{schema}}{\omega}} \in \Phi\\
-    \sevalNA{\stx{schema}~\sttyidx}{\sqity{\qity}}\\
-    \tdeabs{\Phi}{\tc}{\qity}{\memD}{\ity}{\memD'}\\
-    \memD' \leadsto \delta : \Delta\\
-    \Delta \vdash \tau
-}{
-    \tdeabs{\Phi}{\tc}{\srep{\sty{\tc}{\sttyidx}}}{\memD}{\tau}{\memD'}
-}
-\end{mathpar}
-The translation of a user-defined type constructed by any tycon other than the delegated tycon is added to the store (the supplement has the simple rule for retriving it once there):
-\begin{mathpar}
-\small
-\inferrule[abs-tc-foreign-new]{
-    c \neq \tc\\
-    \sty{\tc}{\sttyidx} \notin \text{dom}(\memD)\\
-    \tcdef{\tc}{\psi}{\tcstruct{\stx{schema}}{\omega}} \in \Phi\\\\
-    \sevalNA{\stx{schema}~\sttyidx}{\sqity{\qity}}\\
-    \tdeabs{\Phi}{\tc}{\qity}{\memD}{\ity}{\memD'}\\\\
-    \memD' \leadsto \delta : \Delta\\
-    \Delta \vdash \tau\\
-    (\alpha~\text{fresh})
-}{
-    \tdeabs{\Phi}{c}{\srep{\sty{\tc}{\sttyidx}}}{\memD}{\alpha}{\memD', \sty{\tc}{\sttyidx} \leftrightarrow [\delta]\tau/\alpha}
-}
-\end{mathpar}
-The translation of an ``other'' type is given directly in its index (rule (abs-other-foreign-new) is in the supplement):
-\begin{mathpar}
-\small
-\inferrule[abs-other-local]{
-    \tdeabs{\Phi}{\keyw{other}[m]}{\qity}{\memD}{\tau}{\memD'}
-}{
-    \tdeabs{\Phi}{\keyw{other}[m]}{\srep{\sty{\keyw{other}[m]}{(\stx{nat}, \sqity{\qity}}}}{\memD}{\tau}{\memD'}
-}
-\end{mathpar}
-
-\subsection{Typing Contexts}
-We must also define a translation judgement for external typing contexts $\vdash_\Phi \Upsilon \leadsto \Gamma$ that checks that $\Upsilon$ contains valid types, then generates the corresponding internal typing context $\Gamma$ from their translations (see supplement).
-%\vspace{-5px}
- %We have that external typing contexts obey the standard structural congruences: weakening, exchange and contraction.
-\section{Typing and Translation of External Terms}\label{external-terms}
-
-Having established how types are constructed, and how they determine abstract and from there concrete translations to internal types, we can finally give the typing and translation rules for external terms, shown in Figure \ref{typing}.
-
-Because we are defining a bidirectional type system, a subsumption rule is needed to allow synthetic terms to be analyzed against an equivalent type.
-Per Sec.
-\ref{sec:type-equivalence}, equivalent types must be  syntactically identical at normal form, and we consider analysis only if $\istype{\st}{\Phi}$, so the rule (subsume) is straightforward.
-To use an analytic term in a synthetic position, the programmer must provide a type ascription, written $e : \st$.
-The ascription is kind checked and normalized to a type before being used for analysis by rule (ascribe).
-
-Variables and functions behave in the standard manner given our definitions of types and type translations (used to generate ascriptions in the IL, which is intrinsically rather than bidirectionally typed).
-We use Plotkin's fixpoint operator for general recursion (cf.
-\cite{pfpl}), and define both lambdas and fixpoints only in analytic positions for simplicity.
-
-\subsection{Generalized Introductory Operations}\label{sec:introop}
-The translation of the generalized introductory form, $\eintro{\sttmidx}{\es}$, is determined by the tycon of the type it is being analyzed against as a function of the type's index, the \emph{term index}, $\sttmidx$, and the \emph{argument list}, $\es$.
-
-Before discussing rules (ana-intro) and (ana-intro-other), we note that we can recover a variety of standard concrete introductory forms by a purely syntactic desugaring to this abstract form (and thus allow their use at more than one type).
-For example, for regular strings we can use the string literal form, $\texttt{"s"}$, which desugars to $\eintro{\texttt{"s"}_\text{SL}}{\cdot}$.
-The term index is the corresponding static value of kind $\keyw{Str}$, indicated by a subscript for clarity.
-Similarly, for labeled products, records, objects and so on, we can define a generalized labeled collection form, $\{\mathtt{lbl}_1=e_1, \ldots, \mathtt{lbl}_n=e_n\}$, that desugars to $\eintro{\texttt{[}\mathtt{lbl}_1, \ldots, \mathtt{lbl}_n\texttt{]}}{e_1; \ldots; e_n}$, i.e.
-a list constructed from the row labels is the term index and the corresponding row values are the arguments.
-In both cases, the term index captures  static portions of the concrete form and the arguments capture all external sub-terms.
-Additional desugarings are shown in the supplement and a technique based on \cite{TSLs} could be introduced to allow tycon providers to define more such desugarings  composably.
-
-\begin{figure*}[t]
-\small\fbox{$\eanaX{e}{\st}{\iota}$}~
-\fbox{$\esynX{e}{\st}{\iota}$}
-\begin{mathpar}
-\inferrule[subsume]{
-    \esynX{e}{\st}{\iota}
-}{
-    \eanaX{e}{\st}{\iota}
-}
-
-\inferrule[ascribe]{
-  \sofkz{\emptyset}{\emptyset}{\Phi}{\st}{\kty}\\
-  \snormT{\st}{\st'}\\\\
-  \eanaX{e}{\st'}{\iota}
-}{
-  \esynX{\easc{e}{\st}}{\st'}{\iota}
-}
-
-\small\inferrule[syn-var]{
-  x \Rightarrow \st \in \Upsilon
-}{
-  \esynX{x}{\st}{x}
-}
-
-\inferrule[ana-fix]{
-  \eana{\Upsilon, x \Rightarrow \st}{\Phi}{e}{\st}{\iota}\\\\
-  \sfinalrepX{\st}{\tau}
-}{
-  \eanaX{\efix{x}{e}}{\st}{\ifix{\tau}{x}{\iota}}
-}
-
-\inferrule[ana-lam]{
-    \eana{\Upsilon, x \Rightarrow \st_1}{\Phi}{e}{\st_2}{\iota}\\\\
-    \sfinalrepX{\st_1}{\tau_1}
-}{
-    \eanaX{\eanalam{x}{e}}{\sty{\rightharpoonup}{(\st_1, \st_2)}}{\ilam{\tau_1}{x}{\iota}}
-}
-
-\inferrule[syn-ap]{
-  \esynX{e_1}{\sty{\rightharpoonup}{(\st_1, \st_2)}}{\iota_1}\\
-  \eanaX{e_2}{\st_2}{\iota_2}
-}{
-  \esynX{\eap{e_1}{e_2}}{\st_2}{\iap{\iota_1}{\iota_2}}
-}
-
-\inferrule[ana-intro]{
-  \tcdef{\tc}{\tcsig{\_}{\chi}}{\tcstruct{\_}{\omega}} \in \Phi\\\\
-  \introsig{\klitidx} \in \chi\\
-  \sofkz{\emptyset}{\emptyset}{\Phi}{\stmidx}{\klitidx}\\\\
-  \keyw{ana~intro}={\stx{def}} \in \omega\\
-  |\es| = n\\
-    \keyw{args}(n)=\stx{args}\\\\
-  \stx{def}~\sttyidx~\stmidx~\stx{args} \Downarrow_{\es;\Upsilon;\Phi} \sqitm{\qitm}\\\\
-  %\snorm{\stx{def}~\sttyidx~\stmidx~\stx{args}}{\emptyset}{\memG_0}{{\rightharpoonup}, \tc; \Upsilon; \Phi}{\sqitm{\itm_\text{abs}}}{\memD}{\memG}\\\\
-  %\snorm{\srep{\sty{\tc}{\sttyidx}}}{\memD}{\cdot}{{\rightharpoonup}, \tc;\emptyset;\Phi}{\sqity{\ity_\text{abs}}}{\memD'}{\cdot}\\\\
-  \edeabs{\tc}{{\es;\Upsilon;\Phi}}{\qitm}{\emptyset}{\emptyset}{\iota_\text{abs}}{\memD}{\memG}\\
-  \memG \leadsto \gamma : \Gamma_\text{abs}\\\\
-  \tdeabs{\Phi}{\tc}{\srep{\sty{\tc}{\sttyidx}}}{\memD}{\ity_\text{abs}}{\memD'}\\
-  \memD' \leadsto \delta : \Delta_\text{abs}\\\\
-  \Delta_\text{abs}~\Gamma_\text{abs} \vdash \iota_\text{abs} : \tau_\text{abs}
-}{
-  \eanaX{\eintro{\stmidx}{\es}}{\sty{\tc}{\sttyidx}}{[{\delta}][{\gamma}]\iota}
-}
-
-\inferrule[syn-targ]{
-  \esynX{e_\text{targ}}{\sty{\tc}{\sttyidx}}{\iota_\text{targ}}\\\\
-  \tcdef{\tc}{\tcsig{\_}{\chi}}{\tcstruct{\_}{\omega}} \in \Phi\\\\
-  \opsig{\opname{op}}{\klitidx} \in \chi\\
-  \sofkz{\emptyset}{\emptyset}{\Phi}{\stmidx}{\klitidx}\\\\
-  \keyw{syn~}\opname{op}={\stx{def}} \in \omega\\
-  |e_\text{targ}; \es| = n\\
-    \keyw{args}(n)=\stx{args}\\\\
-  \stx{def}~\sttyidx~\stmidx~\stx{args} \Downarrow_{(e_\text{targ}; \es);\Upsilon;\Phi} (\st, \sqitm{\qitm})\\\\
-  %\snorm{\stx{def}~\sttyidx~\stmidx~\stx{args}}{\emptyset}{\memG_0}{{\rightharpoonup}, \tc; \Upsilon; \Phi}{\sqitm{\itm_\text{abs}}}{\memD}{\memG}\\\\
-  %\snorm{\srep{\sty{\tc}{\sttyidx}}}{\memD}{\cdot}{{\rightharpoonup}, \tc;\emptyset;\Phi}{\sqity{\ity_\text{abs}}}{\memD'}{\cdot}\\\\
-  \edeabs{\tc}{{(e_\text{targ}; \es);\Upsilon;\Phi}}{\qitm}{\emptyset}{\emptyset}{\iota_\text{abs}}{\memD}{\memG}\\
-  \memG \leadsto \gamma : \Gamma_\text{abs}\\\\
-  \tdeabs{\Phi}{\tc}{\srep{\st}}{\memD}{\ity_\text{abs}}{\memD'}\\
-  \memD' \leadsto \delta : \Delta_\text{abs}\\\\
-  \Delta_\text{abs}~\Gamma_\text{abs} \vdash \iota_\text{abs} : \tau_\text{abs}
-}{
-    \esynX{\etarg{\opname{op}}{\sttmidx}{e_\text{targ}}{\es}}{\st}{[\delta][\gamma]\iota_\text{abs}}
-}
-
-\inferrule[ana-intro-other]{
-    \sofkz{\emptyset}{\emptyset}{\Phi}{\stx{def}}{\kargs \rightarrow \kitm}\\\\
-    |\es| = n\\
-    \keyw{args}(n)=\stx{args}\\
-    \stx{def}~\stx{args} \Downarrow_{\es; \Upsilon; \Phi} \sqitm{\qitm}\\\\
-    \edeabs{\keyw{other}[m]}{\es; \Upsilon;\Phi}{\qitm}{\emptyset}{\emptyset}{\itm_\text{abs}}{\memD}{\memG}\\\\
-    \tdeabs{\Phi}{\keyw{other}[m]}{\srep{\sty{\keyw{other}[m]}{\sttyidx}}}{\memD}{\tau_\text{abs}}{\memD'}\\\\
-    \memD' \leadsto \delta : \Delta_\text{abs}\\
-    \memG \leadsto \gamma : \Gamma_\text{abs}\\
-    \Delta_\text{abs}~\Gamma_\text{abs} \vdash \iota_\text{abs} : \tau_\text{abs}
-}{
-    \eanaX{\eintro{\stx{def}}{\es}}{\sty{\keyw{other}[m]}{\sttyidx}}{[\delta][\gamma]\iota_\text{abs}}
-}
-~~~~~~
-\inferrule[syn-targ-other]{
-\esynX{e_\text{targ}}{\sty{\keyw{other}[m]}{\sttyidx}}{\iota_\text{targ}}\\\\
-  \sofkz{\emptyset}{\emptyset}{\Phi}{\stx{def}}{\kargs\rightarrow(\kty\times\kitm)}\\\\
-  |e_\text{targ}; \es| = n\\
-    \keyw{args}(n)=\stx{args}\\
-  \stx{def}~\stx{args} \Downarrow_{(e_\text{targ}; \es);\Upsilon;\Phi} (\st, \sqitm{\qitm})\\\\
-  %\snorm{\stx{def}~\sttyidx~\stmidx~\stx{args}}{\emptyset}{\memG_0}{{\rightharpoonup}, \tc; \Upsilon; \Phi}{\sqitm{\itm_\text{abs}}}{\memD}{\memG}\\\\
-  %\snorm{\srep{\sty{\tc}{\sttyidx}}}{\memD}{\cdot}{{\rightharpoonup}, \tc;\emptyset;\Phi}{\sqity{\ity_\text{abs}}}{\memD'}{\cdot}\\\\
-  \edeabs{\keyw{other}[m]}{{(e_\text{targ}; \es);\Upsilon;\Phi}}{\qitm}{\emptyset}{\emptyset}{\iota_\text{abs}}{\memD}{\memG}\\\\
-  \tdeabs{\Phi}{\keyw{other}[m]}{\srep{\st}}{\memD}{\ity_\text{abs}}{\memD'}\\\\
-  \memD' \leadsto \delta : \Delta_\text{abs}\\
-  \memG \leadsto \gamma : \Gamma_\text{abs}\\
-  \Delta_\text{abs}~\Gamma_\text{abs} \vdash \iota_\text{abs} : \tau_\text{abs}
-}{
-     \esynX{\etarg{\opname{op}}{\stx{def}}{e_\text{targ}}{\es}}{\st}{[\delta][\gamma]\iota_\text{abs}}
-}\vspace{-5px}
-\end{mathpar}
-\caption{Typing}
-\label{typing}
-\end{figure*}
-\begin{figure*}[t]
-\small\fbox{$\vdash_\Phi \omega \sim \psi$}
-\vspace{-25px}
-\begin{mathpar}\small
-\hspace{50px}\inferrule[ocstruct-intro]{
-    \introsig{\klitidx} \in \chi\\
-    \emptyset \vdash \klitidx\\\\
-    \sofkz{\emptyset}{\emptyset}{\Phi}{\stx{def}}{\ktyidx \rightarrow \klitidx \rightarrow \kargs \rightarrow \kitm}
-}{
-    \vdash_\Phi \tcstructn{\stx{def}} \sim \tcsig{\ktyidx}{\chi}
-}
-
-\inferrule[ocstruct-targ]{
-    \vdash_\Phi \omega \sim \tcsig{\ktyidx}{\chi}\\
-    \opname{op}\notin\text{dom}(\chi)\\
-    \emptyset \vdash \klitidx\\\\
-    \sofkn{\emptyset}{\emptyset}{\Phi}{0}{\stx{def}}{\ktyidx \rightarrow \klitidx \rightarrow \kargs \rightarrow (\kty \times \kitm)}
-}{
-    \vdash_\Phi \tcstructc{\omega}{op}{\stx{def}} \sim \tcsig{\ktyidx}{\chi, \opsig{\opname{op}}{\klitidx}}
-}\vspace{-5px}
-\end{mathpar}
-\caption{Checking opcon structures against tycon signatures.}
-\label{ocstruct}\vspace{-5px}
-\end{figure*}
-Let us derive $\small\eana{\Upsilon_\text{ex}}{\Phi_\text{rstr}\Phi_\text{lprod}}{e_\text{ex}}{\stx{paper}}{\iota_\text{ex}}$ where $\small\Upsilon_\text{ex}  := title \Rightarrow \stx{title}$ and $\small e_\text{ex} := \{\conclbl{title}=title, \conclbl{conf}=\texttt{"EXMPL 2015"}\}$.
-The  translation will be $\small\iota_\text{ex} := (title, (\texttt{"EXMPL 2015"}_\text{IL}, ()))$, where $\small\texttt{"EXMPL 2015"}_\text{IL}$ is an internal string (of internal type $\keyw{str}$).
-
-The first premise of (ana-intro) extracts the tycon definition for the tycon of the type the intro form is being analyzed against.
-In this example, this is $\tcvar{lprod}$.
-We will use this as the \emph{delegated tycon} in the final premises of the rule.
-
-The second premise extracts the \emph{intro term index kind}, $\klitidx$, from the \emph{opcon signature}, $\chi$, and the third premise checks the provided term index against this kind.
-This is simply the kind of term index expected by the tycon, e.g., $\tcvar{lprod}$ specifies $\klist{\klbl}$, so that it can use the labeled collection form, while $\tcvar{rstr}$ specifies an intro index kind of $\keyw{Str}$, so that it can use the string literal form.
-%\[\small\begin{array}{lcl}
-%\chi_\text{lprod} & := & \introsig{\klist{\klbl}}, \chi_\text{lprod/targops}\\
-%\chi_\text{rstr} & := & \introsig{\keyw{Str}}, \chi_\text{rstr/targops}
-%\end{array}
-%\]
-
-The fourth premise extracts the \emph{intro opcon definition} from the \emph{opcon structure}, $\omega$, of the tycon structure, calling it $\stx{def}$.
-This is a static function that is applied, in the seventh premise, to determine whether the term is well-typed, raising an error if not or determining the translation of the term if so.
-The function has access to the type index, the term index and an interface to the list of arguments, and its kind is checked by the  judgement $\vdash_\Phi \omega \sim \psi$, which appeared as the final premise of the rule (tcc-ext) and is defined in Figure \ref{ocstruct}.
-
-For example, the opcon structure of $\tcvar{rstr}$ is  $\small\omega_\text{rstr} := \tcstructn{\stx{rstr/intro}}; \omega_\text{rstr/targops}$ where $\small\stx{rstr/intro}:=$
-\[\small\begin{array}{l}
-    \slam{\krx}{\svar{tyidx}}{\slam{\kstr}{\svar{tmidx}}{\slam{\kargs}{\svar{args}}{\\
-\quad \keyw{let}~\svar{aok} :: \kunit = \svar{arity0}~\svar{args}~\keyw{in}~\\
-\quad \keyw{let}~\svar{rok} :: \kunit = \svar{rmatch}~\svar{tyidx}~\svar{tmidx}~\keyw{in}~\svar{str\_of\_Str}~\svar{tmidx}
-}}}
-\end{array}\]
-Because regular strings are implemented as strings, this intro opcon definition is straightforward.
-It begins by making sure that no arguments were passed in (we will return to arguments and the fifth and sixth premises of (ana-intro) with the next example), using the helper function $\svar{arity0} :: \kargs \rightarrow \kunit$ defined such that any non-empty list will raise an error, via the static term $\sraise{\kunit}$.
-In practice, the tycon provider would specify an error message here.
-Next, it checks the string provided as the term index against the regular expression given as the type index using $\svar{rmatch} :: \krx \rightarrow \keyw{Str} \rightarrow \kunit$, which we assume is defined in the usual way and again raises an error on failure.
-Finally, a {translation} corresponding to the term index is generated via  helper function $\svar{str\_of\_Str} :: \keyw{Str} \rightarrow \kitm$.% and $\svar{nat\_of\_Nat} :: \keyw{Nat} \rightarrow \kitm$ to generate  translational internal terms corresponding to the provided static terms and $\svar{len} :: \keyw{Str} \rightarrow \keyw{Nat}$ to statically compute the length of the string.
-
-The only introductory form for kind $\kitm$ is $\sqitm{\qitm}$, where $\qitm$ is a \emph{translational internal term}.
-This form is analagous to the introductory form for kind $\kity$ described in Sec.
-\ref{sec:type-translations}, $\sqity{\qity}$.
-Each form in the syntax of $\iota$ has a corresponding form in the syntax for $\qitm$ and both the kinding rules and dynamics simply recurse through these in the same manner as shown in Sec.
-\ref{sec:type-translations}.
-There is also an analagous unquote form, $\quq{\st}$.
-%The supplement gives the analagous rules.
-
-The final two forms of translational internal term are $\anatrans{n}{\st}$ and $\syntrans{n}$.
-These stand in for the translation of argument $n$, the first if it arises via analysis against type $\st$ and the second if it arises via type synthesis.
-Before giving the rules, let us motivate the mechanism with the intro opcon definition for $\tcvar{lprod}$.
-We have $\omega_\text{lprod} := \tcstructn{\stx{lprod/intro}}; \omega_\text{lprod/targops}$ where $\stx{lprod/intro} :=$\vspace{-5px}
-\[\small
-\begin{array}{l}
-\lambda\svar{tyidx}{:}\klist{\klbl\times\kty}.\lambda\svar{tmidx}{:}\klist{\klbl}.\lambda\svar{args}{:}\kargs.\\
-\quad \keyw{let}~\svar{inhabited}:\kunit=\svar{uniqmap}~\svar{tyidx}\\
-\quad \skap{\kitm}{\skap{\karg}{\skap{\klbl}{\skap{\klbl\times\kty}{\svar{listrec3}}}}}~\svar{tyidx}~\svar{tmidx}~\svar{args}~\sqitm{\itriv}\\
-\quad\quad \lambda \svar{rowtyidx}{:}\klbl\times\kty.\lambda \svar{rowtmidx}{:}\klbl.\lambda\svar{rowarg}{:}\karg.\lambda \svar{r}{:}\kitm.\\
-\quad\quad\quad \keyw{letpair}~(\svar{rowlbl}, \svar{rowty})=\svar{rowtyidx}\\
-\quad\quad\quad \keyw{let}~\svar{lok}::\kunit=\svar{lbleq}~\svar{rowlbl} ~\svar{rowtmidx}\\
-\quad\quad\quad \keyw{let}~\svar{rowtr} :: \kitm = \svar{ana}~\svar{rowarg}~\svar{rowty}\\
-\quad\quad\quad \sqitm{(\quq{\svar{rowtr}}, \quq{\svar{r}})
-}
-\end{array}
-\]
-The first line checks that the type provided is inhabited, in this case by checking that there are no duplicate labels via the helper function $\svar{uniqmap} :: \klist{\klbl \times \kty}\rightarrow \kunit$, raising an error if there are.
-An alternative strategy may have been to use an abstract kind that ensured that such type indices could not have been constructed, but to be compatible with our equality kind restriction, this would require support for abstract equality kinds, analagous to abstract equality types in SML.
-We chose not to formalize these for simplicity, and to demonstrate this general technique.
-An analagous technique could be used to implement record types by requiring that the index be sorted.%, leaving indices where the labels did not appear sorted uninhabited.
-In both cases, we could provide a static helper function of kind $\klist{\klbl\times\kty}\rightarrow \kty$ that checked well-formedness immediately before constructing the requested type.
-
-The rest of this opcon definition folds over the three lists provided as input: the list mapping labels to types provided as the type index, the list of labels provided as the term index, and the list of argument interfaces.
-We assume a straightforward helper function, $\svar{listfold3}$, that raises an error if the three lists are not of the same length.
-The base case is the translational empty product.
-The recursive case first checks that the label provided in the term index matches the label provided in the type index, using a helper function $\svar{lbleq} :: \klbl \rightarrow \klbl \rightarrow \kunit$.
-Then, we request type analysis of the corresponding argument, $\svar{rowarg}$,  against the type in the type index, $\svar{rowty}$, by writing $\svar{ana}~\svar{rowarg}~\svar{rowty}$.
-Here, $\svar{ana}$ is a helper function defined below that produces a translational internal term  of the form $\sqitm{\anatrans{n}{\st}}$, where $\st$ is the value of $\svar{rowty}$, if the row value is well-typed, and raises an error if not.
-The final line constructs a nested tuple based on this translation and the recursive result.
-Taken together, the translational internal term that will be derived for our example involving $e_\text{ex}$ above is $\small\qitm_\text{ex} := (\anatrans{0}{\stx{title}}, (\anatrans{1}{\stx{conf}}, ()))$.
-
-\paragraph{Argument Interface Lists} We define $\karg$, the kind of \emph{argument interfaces}, as a simple product of functions, $\karg := (\kty \rightarrow \kitm) \times (\kunit \rightarrow \kty\times\kitm)$, one for analysis and the other for synthesis, both having the expected kind for these operations.
-The helper functions $\svar{ana}$ and $\svar{syn}$ simply project the corresponding function out, $\svar{ana} := \slam{\karg}{\svar{arg}}{\keyw{fst}(\svar{arg})}$ and $\svar{syn}  :=  \slam{\karg}{\svar{arg}}{\keyw{snd}(\svar{arg})}$.
-
-The argument interface list, called $\stx{args}$ in the rules, is simply a static list of kind $\kargs$ where the $i$th entry is $(\slam{\kty}{\svar{ty}}{\sana{i}{\svar{ty}}}, \slam{\kunit}{\_}{\ssyn{i}})$.
-It is generated by the judgement $\keyw{args}(n)=\stx{args}$, where $n$ is given as the length of the argument list, written $|\es|=n$ (see supplement).
-
-%The $\kargs$ given where the $n$th entry is simply a pair of functions, the first of which allows for analysis of the $n$th argument of the operation against a type, and the second of which requests type synthesis for the $n$th argument.
-The helper functions simply project out the appropriate functions and invoke them.
-
- Recall that the kinding judgement is indexed by $n$, which is an upper bound on the argument index of terms of the form $\sana{n}{\st}$ and $\ssyn{n}$. This is enforced in their kinding rules:
-\begin{mathpar}\small
-\inferrule[k-ana]{
-    n' < n\\
-    \sofkX{\st}{\kty}
-}{
-    \sofkX{\sana{n'}{\st}}{\kitm}
-}
-
-\inferrule[k-syn]{
-    n' < n
-}{
-    \sofkX{\ssyn{n'}}{\kty\times\kitm}
-}
-\end{mathpar}
-
-The following lemma characterizes thus characterizes the argument list interface generation judgement:
-\begin{lemma}
-\small If $\keyw{args}(n)= \stx{args}$ then $\sofkn{\emptyset}{\emptyset}{\Phi}{n}{\stx{args}}{\kargs}$.
-\end{lemma}
-
-The rule (ocstruct-intro) ruled out writing either of these forms explicitly in an opcon definition by checking against the bound $n=0$.
-This is to prevent out-of-bounds errors -- tycon providers do not write these forms directly, but only access them via the argument interface list, guaranteed to have the correct length.
-This is  safe because:
-\begin{lemma}
-\small If $\sofkn{\kDelta}{\kGamma}{\Phi}{n'}{\st}{\kappa}$ and $n > n'
-$ then $\sofkn{\kDelta}{\kGamma}{\Phi}{n}{\st}{\kappa}$.
-\end{lemma}
-
-The dynamics for these forms couples the dynamics of the static language to the statics of the external language.
-For $\sana{n}{\st}$, after stepping the type $\st$ to a value and propagating errors, the argument environment, which stores the arguments themselves and the typing and tycon contexts, $\es; \Upsilon; \Phi$, is consulted to retrieve the $n$th argument and analyze it against $\st$.
-If this succeeds, the translational internal term $\sqitm{\anatrans{n}{\st}}$ is generated to refer to it.
-\begin{mathpar}\small
-\inferrule[s-ana-success]{
-    \svalNA{\st}\\
-    \keyw{nth}[n](\es) = e\\
-    \eana{\Upsilon}{\Phi}{e}{\st}{\iota}
-}{
-    \sstep{\sana{n}{\st}}{\es;\Upsilon;\Phi}{\sqitm{\anatrans{n}{\st}}}
-}
-\end{mathpar}
-If it fails, an error is raised:
-\begin{mathpar}\small
-\inferrule[s-ana-fail]{
-    \svalNA{\st}\\
-    \keyw{nth}[n](\es) = e\\
-    [\Upsilon \vdash_\Phi e \nLeftarrow \st]
-}{
-    \serr{\sana{n}{\st}}{\es;\Upsilon;\Phi}
-}
-\end{mathpar}
-We write $\small[\Upsilon \vdash_\Phi e \nLeftarrow \st]$ to indicate that $e$ fails to analyze against $\st$.
-We do not define this  inductively, so we also allow that this premise be omitted, leaving a non-deterministic semantics nevertheless sufficient for our metatheory.
-
-The dynamics for $\ssyn{n}$ are analagous, evaluating to a pair $(\st, \small\sqitm{\syntrans{n}})$ where $\st$ is the synthesized type.
-
-The kinding rules also prevent the translational internal term forms generated by these static operations from being well-kinded when $n = 0$.
-Like the translational internal type form $\srep{\st}$, these  are retained in values of kind $\kitm$:
-\begin{mathpar}\small
-\inferrule[s-itm-anatrans-v]{
-    \sval{\st}{\es;\Upsilon;\Phi}\\
-    \keyw{nth}[n](\es)=e
-}{
-    \sval{\sqitm{\anatrans{n}{\st}}}{\es;\Upsilon;\Phi}
-}
-
-\inferrule[s-itm-syntrans-v]{
-    \keyw{nth}[n](\es)=e
-}{
-    \sval{\sqitm{\syntrans{n}}}{\es;\Upsilon;\Phi}
-}
-\end{mathpar}
-
-\paragraph{Abstracted Term Translations} The reason for this  is again because we will hold these abstract by replacing them with variables.
-The judgement $\edeabs{\tc}{\argEnv}{\qitm}{\memD}{\memG}{\itm\moutput}{\memD'\moutput}{\memG'\moutput}$, appearing as the eighth premise of (ana-intro), relates a translational internal term $\qitm$ to an internal term $\itm$ called the corresponding \emph{abstracted term translation}, where all references to the translation of an argument (having any type) are replaced with a unique variable and all references to the translation of a type constructed by a user-defined tycon other than the ``delegated tycon'' $\tc$ are replaced with an abstract type variable as described in Sec.
-\ref{sec:selective-type-translation-abstraction}.
-The type translation store, $\memD$, discussed previously, and term translation store, $\memG$, track these mappings.
-Term translation stores have the following syntax: $\memG ::= \emptyset ~|~ \memG, n : \st \leadsto \iota/x : \tau$.
-Each entry can be read ``argument $n$ having type $\st$ and translation $\iota$ appears as variable $x$ with type $\tau$''.
-For example, \[\small\edeabs{\tcvar{lprod}}{(title, \texttt{"EXMPL 2015"}),\Upsilon,\Phi_\text{rstr}\Phi_\text{lprod}}{\qitm_\text{ex}}{\emptyset}{\emptyset}{\iota_\text{ex/abs}}{\memD_\text{ex/abs}}{\memG_\text{ex/abs}}\] where $\itm_\text{ex/abs} := (x_0, (x_1, ()))$, the term translation store is $\memG_\text{ex/abs} := 0 : \stx{title} \leadsto title/x_0 : \alpha_0, 1 : \stx{conf} \leadsto \texttt{"EXMPL 2015"}_\text{IL}/x_1 : \alpha_1$ and the type translation store $\memD_\text{ex/abs}$ is alpha-equivalent to $\memD_\text{paper/abs}$ from Sec.
-\ref{sec:selective-type-translation-abstraction}.
-
-The rules for the abstracted term translation judgement follow recursively for shared forms just like those for the abstracted type translation judgement (see supplement).
-The rule for references to argument translations derived via type analysis operates as follows ($\syntrans{n}$ is analagous):
-\begin{mathpar}\small
-\inferrule[abs-anatrans-new]{
-    n \notin \text{dom}(\memG)\\
-    \keyw{nth}[n](\es) = e\\
-    \eanaX{e}{\st}{\iota}\\
-    \tdeabs{\Phi}{c}{\srep{\st}}{\memD}{\ity}{\memD'}\\
-    (x~\text{fresh})
-}{
-    \edeabs{c}{\es;\Upsilon;\Phi}{\anatrans{n}{\st}}{\memD}{\memG}{x}{\memD'}{\memG, n : \st \leadsto \iota/x : \tau}
-}
-\end{mathpar}
-
-Like $\memD$, each $\memG$ induces an internal term substitution,  $\gamma ::= \emptyset ~|~ \gamma, \iota/x$, and corresponding internal typing context $\Gamma$ by the judgement $\memG \leadsto \gamma : \Gamma$, appearing as the ninth premise.
-In this case, $\small
-\gamma_\text{ex/abs} := title/x_0, \texttt{"EXMPL 2015"}_\text{IL}/x_1$ and $\small\Gamma_\text{ex/abs} := x_0 : \alpha_0, x_1 : \alpha_1$.
-
-The tenth premise of (ana-intro) determines an abstract type translation for the type provided for analysis, in this case $\small\tau_\text{ex/abs} := \alpha_0 \times (\alpha_1 \times \kunit)$ (alpha-equivalent to $\tau_\text{paper/abs}$ in Sec.
-\ref{sec:selective-type-translation-abstraction}), and the eleventh premise extracts a type substitution, $\delta_\text{ex/abs}$, and type formation context, $\Delta_\text{ex/abs}$, from $\memD_\text{ex/abs}$, again equivalent to $\delta_\text{conf/abs}$ and $\Delta_\text{conf/abs}$ from Sec.
-\ref{sec:selective-type-translation-abstraction}.
-
-Finally, the twelfth premise checks the abstracted term translation against the abstracted type translation.
-Here, we are checking $\Delta_\text{ex/abs}~\Gamma_\text{ex/abs} \vdash \iota_\text{ex/abs} : \tau_\text{ex/abs}$, i.e.: \[\small(\alpha_0, \alpha_1)~(x_0 : \alpha_0, x_1 : \alpha_1) \vdash (x_0, (x_1, ())) : \alpha_0 \times (\alpha_1 \times \kunit)
-\]
-
-In other words, the translation of the labeled product $e_\text{ex}$ generated by $\tcvar{lprod}$ is checked with all references to term and type translations of regular strings replaced by variables and type variables, respectively.
-Nevertheless, because our intro opcon definition treated arguments parametrically, the check succeeds.
-We will describe  an ill-behaved operator, which would lead to a violation of the regular string tycon  invariant were it not disallowed by this check,  in Sec.
-\ref{metatheory}.
-
-Applying the substitutions $\gamma_\text{ex/abs}$ and $\delta_\text{ex/abs}$ in the conclusion of the rule, we arrive at $\iota_\text{ex}$.
-
-\paragraph{Other Intro} The rule (ana-intro-other) is used to introduce terms of a type constructed by an ``other'' tycon.
-The term index, rather than the tycon context, directly specifies the static function that maps the arguments to a translation.
-In all other respects, the rule is analagous.
-It is used as a technical device in our proof of conservativity in Sec.
-\ref{metatheory}.
-
-\subsection{Generalized Targeted Operations} 
-All non-introductory opcons associated with user-defined tycons go through another generalized form, in this case for \emph{targeted operations}, $\etarg{\opname{op}}{\sttmidx}{e_\text{targ}}{\es}$, where $\opname{op}$ ranges over opcon names, ${\sttmidx}$ is the term index, $e_\text{targ}$ is the \emph{target argument} and $\es$ are the remaining arguments.
-
-Desugarings include an explicit form, $e_\text{targ}.\opname{op}\langle\sttmidx\rangle(\es)$ (and variants of where the term index or the arguments are omitted), projection syntax for use by record-like types, $e_\text{targ}\opname{\#}\conclbl{lbl}$, which desugars to $\etarg{\opname{prj}}{\conclbl{lbl}}{e_\text{targ}}{\cdot}$, and concatenative syntax $e_\text{targ} \cdot e_\text{arg}$, which desugars to $\etarg{\opname{conc}}{\striv}{e_\text{targ}}{e_\text{arg}}$.
-We show other desugarings, including case analysis, in the supplement.
-
-Whereas introductory operations were analytic, targeted operations are synthetic in @$\lambda$.
-The type and translation are determined by the tycon of the type synthesized by the target argument.
-The rule (syn-targ) is otherwise similar to (ana-intro) in its structure.
-The first premise synthesizes a type, $\sty{\tc}{\sttyidx}$, for the target argument.
-The second premise extracts the tycon definition for $\tc$ from the tycon context.
-The third extracts the \emph{operator index kind} from its opcon signature, and the fourth checks the term index against this kind.
-For example, we may associate the following opcon signatures with the tycons $\tcvar{rstr}$ and $\tcvar{lprod}$:\[\small
-\begin{array}{lcl}
-\chi_\text{rstr} & := & \keyw{intro}[\keyw{Str}], \opsig{\opname{conc}}{\kunit}, \opsig{\opname{case}}{\klist{\keyw{StrPattern}}},  \\
-& & \opsig{\opname{coerce}}{\krx}, \opsig{\opname{check}}{\krx}, \opsig{\opname{replace}}{\krx}\\
-\chi_\text{lprod} & := & \keyw{intro}[\klist{\klbl}], \opsig{\opname{prj}}{\klbl}, \opsig{\opname{conc}}{\kunit}, \opsig{\opname{drop}}{\klist{\klbl}}
-\end{array}
-\]
-
-The opcons associated with $\tcvar{rstr}$ are taken directly from Fulton et al.'s specification of regular string types \cite{sanitation-psp14}, with the exception of $\opname{case}$, which generalizes case analysis as defined there to arbitrary string patterns, which we discuss in the supplement.
-The opcons associated with $\tcvar{lprod}$ are also straightforward: $\opname{prj}$ projects out the row with the provided label, $\opname{conc}$ concatenates two labeled products (updating common fields with the value of the right argument), and $\opname{drop}$ creates a new labeled product from the target with some fields dropped.
-Note that both regular strings and labeled products can define concatenation without conflict.
-%Targeted operations use no new mechanisms, so we only show regular string concatenation; others are in the supplement.
-
-The fifth premise of (syn-targ) extracts the targeted opcon definition of $\opname{op}$ from the opcon structure, $\omega$.
-Like the intro opcon definition, this is a static function that generates a translational internal term on the basis of the target tycon's type index, the term index and an argument interface list.
-Targeted opcon definitions additionally synthesize a type.
-The rule (ocstruct-targ) in Figure \ref{ocstruct} ensures that no tycon defines an opcon twice and that the opcon definitions are well-kinded.
-For example, $\omega_\text{rstr/targops}$ defines:
-\[\small
-\begin{array}{l}
-\keyw{syn}~\opname{conc}=\slam{\krx}{\svar{tyidx}}{
-    \slam{\kunit}{\svar{tmidx}}{
-        \slam{\kargs}{\svar{args}}{\\
-            \quad \keyw{letpair}~(\svar{arg1}, \svar{arg2}) = \svar{arity2}~\svar{args}\\
-            \quad \keyw{letpair}~(\_, \svar{tr1}) = \svar{syn}~\svar{arg1}\\
-            \quad \keyw{letpair}~(\svar{ty2}, \svar{tr2}) = \svar{syn}~\svar{arg2}\\
-            \quad \stycase{\tcvar{rstr}}{\svar{ty2}}{\svar{tyidx2}}{\\
-                \quad\quad (\sty{\tcvar{rstr}}{\svar{rxconcat}~\svar{tyidx}~\svar{tyidx2}}, \sqitm{sconcat~\quq{\svar{tr1}}~\quq{\svar{tr2}}})\\
-                \quad
-            }{\sraise{\kty\times\kitm}}
-        }
-    }
-}
-\end{array}
-\]
-The first line checks that exactly two arguments, including the target argument, were passed in via helper function $\svar{arity2}$.
-We then request synthesis of both arguments.
-We can ignore the type synthesized by the first because by definition it is a regular string type with type index $\svar{tyidx}$.
-We case analyze the second against $\tcvar{rstr}$, extracting its index regular expression if so and failing if not.
-We finally synthesize the resulting regular string type, using the helper function $\svar{rxconcat} :: \krx \rightarrow \krx \rightarrow \krx$ which generates the concatenated regular expression, and the translation, using an internal helper function $sconcat : \keyw{str} \rightarrow \keyw{str} \rightarrow \keyw{str}$, the translational internal term for which we assume has been substituted in directly above.
-
-The remaining premises of (syn-targ) are analagous to the corresponding premises in (ana-intro), with the only difference being that we check the abstract term translation against the abstract type translation of the synthesized type.
-Like (ana-intro-other), the rule (syn-targ-other) is used when the target synthesizes an ``other'' type.
-The mapping from the arguments to a type and translation is given directly in the term index (the operator name is ignored).
-
-% \section{Metatheory}\label{metatheory}
-% %This judgement is only defined for values of kind $\kty$. We write $\st~\texttt{type}_\Phi$ iff $\vdash \Phi$ and $\sofkn{\emptyset}{\emptyset}{\Phi}{0}{\st}{\kty}$ and $\sval{\st}$.
-
-% \begin{definition}[Argument Environment Formation] $\argEnvOK{n}{\es;\Upsilon;\Phi}$ iff $|\es|=n$ and $\vdash \Phi$ and $\vdash_\Phi \Upsilon \leadsto \Gamma$.\end{definition}
-
-% \subsection{Static Language}
-
-% %\begin{lemma}
-% %If $\kDelta \vdash \kGamma$ and $\vdash \Phi$ and $\sofkn{\kDelta}{\kGamma}{\Phi}{n}{\st}{\kappa}$ then $\kDelta \vdash \kappa$.\todo{don't think we need this}
-% %\end{lemma}
-
-% \begin{theorem}[Static Canonical Forms] If $\sofkn{\emptyset}{\emptyset}{\Phi}{n}{\st}{\kappa}$ and $\argEnvOK{n}{\argEnv}$ and $\sval{\st}{\argEnv}$ then:
-% \begin{enumerate}
-% \item TODO: arrow
-% \item TODO: unit
-% \item TODO: product
-% \item TODO: sum
-% \item TODO: inductive
-% \item TODO: universal
-% \item If $\kappa = \kty$ then $\istype{\st}{\Phi}$ and 
-%     \begin{enumerate}
-%     \item $\st = \sty{\rightharpoonup}{(\st_1, \st_2)}$ and $\istype{\st_1}{\Phi}$ and $\istype{\st_2}{\Phi}$; or
-%     \item $\st = \sty{\tc}{\sttyidx}$ and $\tcdef{\tc}{\tcsig{\ktyidx}{\chi}}{\omega} \in \Phi$ and $\sofkn{\emptyset}{\emptyset}{\Phi}{0}{\st}{\ktyidx}$ and $\svalNA{\sttyidx}$; or
-%     \item $\st = \sotherty{m}{\tau}$ and $\emptyset \vdash \tau$.
-%     \end{enumerate}
-% \item If $\kappa = \kity$ then $\st=\sqity{\qity}$ and $\sofkn{\emptyset}{\emptyset}{\Phi}{0}{\st}{\kity}$ and $\svalNA{\st}$ and $\qtuq{\st'} \notin \qity$ and 
-%     \begin{enumerate}
-%     \item The outer form of $\qity$ is shared with $\tau$ and if $\qity'$ is a sub-term of $\qity$ then $\sofkn{\emptyset}{\emptyset}{\Phi}{0}{\sqity{\qity'}}{\kity}$ and $\svalNA{\sqity{\qity'}}$; or
-%     \item $\qity = \srep{\st'}$ and $\istype{\st'}{\Phi}$
-%     \end{enumerate}
-% \item If $\kappa = \kitm$ then $\st=\sqitm{\qitm}$ and no sub-terms of $\qitm$ have the form $\quq{\st'}$ and 
-%     \begin{enumerate}
-%     \item The outer form of $\qitm$ is shared with $\itm$ and if $\qitm'$ is a sub-term of $\qitm$ then $\sofkn{\emptyset}{\emptyset}{\Phi}{n}{\sqitm{\qitm'}}{\kitm}$ and $\sval{\sqitm{\qitm'}}{\argEnv}$ and if $\qity$ is a sub-term of $\qitm$ then $\sofkn{\emptyset}{\emptyset}{\Phi}{0}{\sqity{\qity}}{\kity}$ and $\svalNA{\sqity{\qity}}$; or
-%     \item $\qitm=\anatrans{n'}{\st'}$ and $n' < n$ and $\istype{\st'}{\Phi}$; or
-%     \item $\qitm=\syntrans{n'}$ and $n' < n$.
-%     \end{enumerate}
-% \end{enumerate}
-% \end{theorem}
-
-% \begin{theorem}[Static Progress]
-% If $\sofkn{\emptyset}{\emptyset}{\Phi}{n}{\st}{\kappa}$ and $\vdash \Phi$ and $\vdash^n \argEnv$ then $\sval{\st}{\argEnv}$ or $\serr{\st}{\argEnv}$ or $\sstep{\st}{\argEnv}{\st'}$.
-% \end{theorem}
-% \begin{proof}We proceed by rule induction on the kinding judgement.
-% (k-parr)
-
-% (k-ty)
-
-% (k-otherty)
-
-% (k-tycase-parr)
-
-% (k-tycase)
-
-% (k-ity-lam)
-
-% (k-ity-alpha)
-
-% (k-ity-unquote)
-
-% (k-ity-trans)
-
-% (k-raise)
-
-% (k-itm-var)
-
-% (k-itm-lam)
-
-% (k-itm-unquote)
-
-% (k-itm-anatrans)
-
-% (k-itm-syntrans)
-% \end{proof}
-
-% \begin{theorem}[Static Preservation]
-% If $\sofkn{\emptyset}{\emptyset}{\Phi}{n}{\st}{\kappa}$ and $\vdash \Phi$ and $\vdash^n \argEnv$ and $\sstep{\st}{\argEnv}{\st'}$ then $\sofkn{\emptyset}{\emptyset}{\Phi}{n}{\st'}{\kappa}$.
-% \end{theorem}
-% \begin{proof}
-% (s-ty-step) (2 cases)
-% (s-tycase-step) (2 cases)
-% (s-tycase-match) (2 cases)
-% (s-tycase-fail-1)
-% (s-tycase-fail-2)
-% (s-ity-lam-step-1)
-% (s-ity-lam-step-2)
-% (s-ity-unquote-step)
-% (s-ity-unquote-elim)
-% (s-ity-trans-step)
-% (s-itm-lam-step-1)
-% (s-itm-lam-step-2)
-% (s-itm-unquote-step)
-% (s-itm-unquote-elim)
-% (s-ana-step)
-% (s-ana-success)
-% (s-syn-success) (relies on type synthesis theorem)
-% (s-itm-anatrans-step)
-% \end{proof}
-
-% % \begin{lemma}[Kinding Stability]
-% % If $\vdash \Phi$ and $\vdash \Phi, \tcdef{\tc}{\psi}{\theta}$ and $\kDelta \vdash \kGamma$ and $\sofkn{\kDelta}{\kGamma}{\Phi}{n}{\st}{\kappa}$ then $\sofkn{\kDelta}{\kGamma}{\Phi, \tcdef{\tc}{\psi}{\theta}}{n}{\st}{\kappa}$.
-% % \end{lemma}
-% % \begin{proof}
-% % all cases straightforward by induction
-% % \end{proof}
-
-% % TODO: static normalization stability inside A?
-
-% \subsection{Types}
-% \begin{lemma}[Type Substitution Application]\label{thm:type-substitution-application}
-% If $\vdash \delta : \Delta$ and $\Delta \vdash \tau$ then $\emptyset \vdash [\delta]\tau$.
-% \end{lemma}
-
-% \begin{lemma}[Selective Type Abstraction]\label{thm:selective-type-abstraction}
-% If $\vdash \Phi$ and $\sofkz{\emptyset}{\emptyset}{\Phi}{\sqity{\qity}}{\kity}$ and $\svalNA{\sqity{\qity}}$ and $\memD \leadsto \delta : \Delta$ and $\vdash \delta : \Delta$ and $\tdeabs{\tc}{\Phi}{\qity}{\memD}{\tau}{\memD'}$ then $\memD' \leadsto \delta' : \Delta'$ and $\delta \subseteq \delta'$ and $\Delta \subseteq \Delta'$ and $\vdash \delta' : \Delta'$ and $\Delta' \vdash \tau$.
-% \end{lemma}
-% \begin{proof}
-% (shared forms) (trans(st))
-% \end{proof}
-
-% \begin{lemma}[Type Translation]\label{type-translation}
-% If $\vdash \Phi$ and $\istype{\st}{\Phi}$ and $\vdash_\Phi \st \leadsto \tau$ then $\emptyset \vdash \tau$.
-% \end{lemma}
-% \begin{proof} By Lemma \ref{thm:type-substitution-application} and Lemma \ref{thm:selective-type-abstraction}.
-% \end{proof}
-
-% \begin{lemma}[Typing Context Translation]
-% If $\vdash \Phi$ and $\vdash_\Phi \Upsilon \leadsto \Gamma$ then $\emptyset \vdash \Gamma$.\end{lemma}
-% \begin{proof} We proceed by rule induction on typing context translation.
-Empty case trivial.
-Extended case follows by Lemma \ref{type-translation} and definition of internal typing context formation.\end{proof}
-
-% \subsection{External Terms}
-% \begin{theorem}[Type Synthesis]
-% If $\vdash \Phi$ and $\vdash_\Phi \Upsilon \leadsto \Gamma$ and $\esynX{e}{\st}{\iota}$ then $\istype{\st}{\Phi}$ and $\vdash_\Phi \st \leadsto \tau$. \end{theorem}
-% \begin{proof}
-% (var) (ap) (syn-targ)
-% \end{proof}
-
-% \begin{lemma}[Selective Term Abstraction]
-% If $\vdash \Phi$ and $\vdash^n \argEnv$ and $\sofkn{\emptyset}{\emptyset}{\Phi}{n}{\sqitm{\qitm}}{\kitm}$ and $\sval{\sqitm{\qitm}}{\argEnv}$ and $\memD \leadsto \delta : \Delta$ and $\vdash \delta : \Delta$ and $\emptyset \vdash \Gamma_\text{out}$ and $\memG \leadsto \gamma : \Gamma$ and $\Delta~\Gamma_\text{out} \vdash \gamma : \Gamma$ and $\edeabs{\tc}{\argEnv}{\qitm}{\memD}{\memG}{\itm}{\memD'}{\memG'}$ then $\memD' \leadsto \delta' : \Delta'$ and $\memG' \leadsto \gamma' : \Gamma'$ and $\delta \subseteq \delta'$ and $\Delta \subseteq \Delta'$ and $\gamma \subseteq \gamma'$ and $\Gamma \subseteq \Gamma'$ and $\vdash \delta' : \Delta'$ and $\Delta'~\Gamma_\text{out} \vdash \gamma' : \Gamma'$.
-% \end{lemma}
-% \begin{proof} By rule induction on selective term abstraction judgement.
-
-% (shared forms) (unquote form not possibly by canonical forms) (anatrans) (syntrans)
-% \end{proof}
-
-% \begin{theorem}[Type-Preserving Translation]
-% If $\vdash \Phi$ and $\vdash_\Phi \Upsilon \leadsto \Gamma$ and either
-% \begin{enumerate}
-% \item $\eanaX{e}{\st}{\iota}$ and $\istype{\st}{\Phi}$; or 
-% \item $\esynX{e}{\st}{\iota}$
-% \end{enumerate}
-% then $\vdash_\Phi \st \leadsto \tau$ then $\emptyset~\Gamma \vdash \iota : \tau$.
-% \end{theorem}
-% \begin{proof} By rule induction on typing rules.
-
-% (var) (lam) (ap) (fix) (subsume) (ascribe) (ana-intro) (syn-targ) (other)
-% \end{proof}
-
-% TODO: stability of typing
-% TODO: Unicity of typing
-% TODO: hygiene?
-
-% TODO: Make definitions out of these.
-% The standard judgement $\Gamma \vdash \gamma : \Gamma'$ states that every binding $x : \tau$ in $\Gamma'$ has a corresponding substitution $\iota/x$ in $\gamma$ such that $\Gamma \vdash \iota : x$.
-
-% , and  the judgement $\vdash \delta : \Delta$ checks that every type variable in $\Delta$ has a well-formed substitution in $\delta$
-
-% \subsection{Conservativity}
-% \begin{theorem}[Conservativity]
-% If $\vdash \Phi$ and $\istype{\sty{\tc}{\sttyidx}}{\Phi}$ and for all $e$, if $\eana{\emptyset}{\Phi}{e}{\sty{\tc}{\sttyidx}}{\iota}$ and $\iota \Downarrow \iota'$ then $P(\iota')$, then if $\vdash \Phi, \tcdef{\tc'}{\psi}{\theta}$ then for all $e$, if $\eana{\emptyset}{\Phi, \tcdef{\tc'}{\psi}{\theta}}{e}{\sty{\tc}{\sttyidx}}{\iota}$ and $\iota \Downarrow \iota'$ then $P(\iota')$.
-% \end{theorem}
-
-% \begin{lemma}[Other Substitution]
-% If $\vdash \Phi, \tcdef{\tc}{\psi}{\theta}$ and $\istype{\st}{\Phi}$ and $\eana{\Upsilon}{\Phi, \tcdef{\tc}{\psi}{\theta}}{e}{\st}{\iota}$ then there exists $e'$ such that $\eana{\Upsilon}{\Phi}{e'}{\st}{\iota}$.
-% \end{lemma}
-
-\section{Metatheory}\label{metatheory}
-We will now state the key metatheoretic properties of @$\lambda$. The full proofs are in the supplement.
-
-\paragraph{Kind Safety} Kind safety ensures that normalization of well-kinded static terms cannot go wrong. We can take a standard progress and preservation based approach. 
-\begin{theorem}[Static Progress]
-If $\sofkn{\emptyset}{\emptyset}{\Phi}{n}{\st}{\kappa}$ and $\vdash \Phi$ and $|\es|=n$ and $\vdash_\Phi \Upsilon \leadsto \Gamma$ then $\sval{\st}{\es; \Phi; \Upsilon}$ or $\serr{\st}{\es; \Phi; \Upsilon}$ or $\sstep{\st}{\es; \Phi; \Upsilon}{\st'}$.
-\end{theorem}
-
-\begin{theorem}[Static Preservation]
-If $\sofkn{\emptyset}{\emptyset}{\Phi}{n}{\st}{\kappa}$ and $\vdash \Phi$ and $|\es|=n$ and $\vdash_\Phi \Upsilon \leadsto \Gamma$ and $\sstep{\st}{\es; \Phi; \Upsilon}{\st'}$ then $\sofkn{\emptyset}{\emptyset}{\Phi}{n}{\st'}{\kappa}$.
-\end{theorem}
-The case in the proof for the form $\ssyn{n}$ requires that the following theorem be simultaneously defined.
-The mutual induction is well-founded because the total number of argument lists in the terms being considered decreases.
-\begin{theorem}[Type Synthesis]
-If $\vdash \Phi$ and $\vdash_\Phi \Upsilon \leadsto \Gamma$ and $\esynX{e}{\st}{\iota}$ then $\vdash_\Phi \st \leadsto \tau$ (and thus $\istype{\st}{\Phi}$). 
-\end{theorem}
-\paragraph{Type Safety}
-Type safety in a typed translation semantics requires that well-typed external terms translate to well-typed internal terms.
-Type safety for the IL \cite{pfpl} then implies that well-typed external terms cannot go wrong.
-To prove this, we must prove a stronger theorem, \emph{type-preserving translation}, which is the analog of type-preserving compilation in the typed compilation literature \cite{tarditi+:til-OLD}\todo{right citation for TIL?}:% produced as the translation of the external type.
-Note that we need only to state our top-level theorems for the analysis judgement  because of the subsumption rule.
-
-\begin{theorem}[Type Preserving Translation]
-If $\vdash \Phi$ and $\vdash_\Phi \Upsilon \leadsto \Gamma$ and $\vdash_\Phi \st \leadsto \tau$ and $\eanaX{e}{\st}{\iota}$ then $\emptyset~\Gamma \vdash \iota : \tau$.
-\end{theorem}
-\begin{proof-sketch}
-We induct on the typing judgement.
-The interesting cases are (ana-intro), (ana-intro-other), (syn-trans) and (syn-trans-other); the latter two arise via subsumption.
-The result follows directly from the final premise of each rule, combined with lemmas that states that if $\memD \leadsto \delta : \Delta_\text{abs}$ and $\memG \leadsto \gamma : \Gamma_\text{abs}$ then $\vdash \delta : \Delta$ and $\Delta~\Gamma \vdash \gamma : \Gamma_\text{abs}$, i.e.
-that all the variables in the contexts have well-formed/well-typed substitutions, and thus that applying them in the conclusions of the rules gives a well-typed term.
-\end{proof-sketch}
-
-\paragraph{Hygienic Translation} 
-Note above that the domains of $\Upsilon$ (and thus $\Gamma$)  and $\Gamma_\text{abs}$ are disjoint.
-This serves to ensure \emph{hygienic translation} -- translations cannot refer to variables in the surrounding scope directly, so uniformly renaming a variable cannot change the meaning of a program.
-Variables in $\Upsilon$ can  occur in arguments (e.g.
-$title$ in the previous example), but the translations of the arguments only appear \emph{after} the substitution $\gamma$ has been applied.
-We assume that substitution is capture-avoiding in the usual manner.
-%(i.e.
-consistent with a locally nameless implementation).
+
+\subsection{Equality Kinds}
+Need an equational theory for SL to state equality kind property, but not important for other metatheory.
+
+\subsection{Full Examples}
+
+\section{Static Language}
+\subsection{Kind Formation}
+
+\subsection{Kinding Context Formation}
+
+\subsection{Kinding}
+
+\subsection{Dynamic Semantics}
+
+\subsection{Kind Safety}
+
+\section{Types}
+\subsection{Type Translations}
+
+\subsection{Typing Context Translations}
 
@@
 
-\paragraph{Stability}
-Extending the tycon context does not change the meaning of any terms that were previously well-typed.
-\begin{theorem}[Stability]
-Letting $\Phi' := \Phi, \tcdef{\tc}{\psi}{\theta}$, if $\vdash \Phi'$ and $\vdash_\Phi \Upsilon \leadsto \Gamma$ and $\vdash_\Phi \st \leadsto \tau$ and $\eanaX{e}{\st}{\iota}$ then, $\vdash_{\Phi'} \Upsilon \leadsto \Gamma$ and $\vdash_{\Phi'} \st \leadsto \tau$ and $\eana{\Upsilon}{\Phi'}{e}{\st}{\iota}$.
-\end{theorem}
-
-\paragraph{Conservativity} 
-Extending the tycon context also conserves all \emph{tycon invariants} maintained in any smaller tycon context. An example of a tycon invariant is the following:
-
-\begin{tyconinvariant}[Regular String Soundness]
-If $\eana{\emptyset}{\Phi_\text{rstr}}{e}{\sty{\tcvar{rstr}}{\concrx{r}}}{\iota}$ and $\iota \Downarrow \iota'$ then $\iota'=\texttt{"s"}$ and $\texttt{"s"}$ is in the regular language $\mathcal{L}(\texttt{r})$.
-\end{tyconinvariant}
-\begin{proof-sketch} The proof is not unusually difficult because we have fixed the tycon context $\Phi_\text{rstr}$, so we can simply treat the calculus like a type-directed compiler for a calculus with only two tycons, $\rightharpoonup$ and $\tcvar{rstr}$.
-Such a calculus and compiler specification was given by Fulton et al.
-\cite{sanitation-psp14} and extending this to our  setting requires showing only that the opcon definitions in $\tcvar{rstr}$ adequately capture these specifications using standard program correctness techniques for the SL, which is a simple functional language.
-The only ``wrinkle'' is that the rule (syn-targ-other) can synthesize a regular string type.
-Luckily the abstracted translation will always be checked against $\tau_\text{abs}=\alpha$, so no strings that could not arise by another rule could  be generated, by the standard abstraction theorem for the IL.
-\end{proof-sketch}
-
-The reason why the rule (syn-targ-other) is never a problem in proving a tycon invariant turns out to be the same reason extending the tycon context conserves all tycon invariants.
-A newly introduced tycon defining a targeted operator that synthesizes a regular string type, e.g.
-$\stx{paper}$, and generating a translation that is not in the corresponding regular language, e.g.
-$\texttt{""}$, could be defined, but when used, the rule (syn-targ) would check the translation against $\tau_\text{abs}=\alpha$, so the invariant could not be broken by the abstraction theorem for the IL.
-%Type abstraction is the basis for conservatively composing type system fragments, just like it is the basis for composing ML-style modules.
- We can state this conservativity property more generally:
-
-\begin{theorem}[Conservativity] If $\vdash \Phi$ and $\tc \in \text{dom}(\Phi)$ and a tycon invariant for $\tc$ holds under $\Phi$: \begin{itemize}
-\item If $\eana{\emptyset}{\Phi}{e}{\sty{\tc}{\sttyidx}}{\iota}$ and $\vdash_\Phi {\sty{\tc}{\sttyidx}} \leadsto \tau$ and $\iota \Downarrow \iota'$ then $P(\sttyidx, \iota')$.
-\end{itemize} then for all $\Phi' = \Phi, \tcdef{\tc'}{\psi'}{\theta'}$ such that $\vdash \Phi'$, the same tycon invariant holds under $\Phi'$: \begin{itemize}
-\item If $\eana{\emptyset}{\Phi'}{e}{\sty{\tc}{\sttyidx}}{\iota}$ and $\vdash_{\Phi'}{\sty{\tc}{\sttyidx}}\leadsto{\tau}$ and $\iota \Downarrow \iota'$ then $P(\sttyidx, \iota')$.
-\end{itemize}
-\end{theorem}
-\begin{proof-sketch}
-The proof maps every well-typed term in the extended tycon context to a well-typed term in the original tycon context with the same translation, so we can apply the original invariant to arrive at the property $P(\sttyidx, \iota')$.
-This term is constructed by replacing all types constructed by $\tc'$ with a type constructed by $\keyw{other}[m]$, for an $m$ uniquely associated with $\tc'$, and all introductory and targeted sub-terms associated with $\tc'$ with an equivalent one that passes through the rules (ana-intro-other) and (syn-targ-other) by simply partially applying the intro and targeted opcon definitions to generate the term indices.
-Types constructed by $\keyw{other}[m]$ are indistinguishable from types constructed by a ``future'' tycon because the type case analysis construct is not defined for this tycon form, so the translations coincide.\end{proof-sketch}
-%Had the translational term generated by the intro opcon definition been, e.g., $(\anatrans{0}{\stx{title}}, (\texttt{"TEST"}, ())$, then the corresponding abstract term would be $(x_0, (\texttt{"TEST"}, ())$ and the check would fail because $(\alpha_0, \alpha_1)~(x_0 : \alpha_0) \nvdash \texttt{"TEST"} : \alpha_1$.
-Had we not gone through the machinations of holding types and terms abstract, however, the check would succeed, because $\keyw{str}$ would no longer have been held abstract as $\alpha_1$.
-This form of ``type translation independence'' is precisely analagous to the representation independence properties that underly abstraction theorems for module systems based on  abstract types and will similarly serve to ensure that the type invariants maintained by $\tcvar{rstr}$ are conserved when new tycons are defined.
-In this case the invariant is that only strings that are in the regular language specified in the type index can be generated from an external term of regular string type.
-Note that $\texttt{"TEST"}$ is not in the regular language specified by $\concrx{[A-Z]+ \digit\digit\digit\digit}$, so it is critical that it not be allowed, despite being consistent with the type translation of $\stx{conf}$, i.e.
-it is a string.
-Such an invariant could not be maintained if the type system were treated as merely a ``bag of rules''.\todo{put conservativity here?} %We consider this more rigorously in Sec.
-\ref{metatheory}.
-
-\section{Related Work}\label{prior-work}
-
-% Note also that type constructors are not themselves static values.
-To apply them in a curried manner or pass them around in the static language, we can define a static function, e.g.
-$\slam{\kty}{\svar{t1}}{\slam{\kty}{\svar{t2}}{\sparr{(\svar{t1}, \svar{t2})}}}$ has kind $\karrow{\kty}{\karrow{\kty}{\kty}}$.
-
-%To situate our work, and because we build on some recent results, we begin by summarizing prior approaches that can be considered when a simple isomorphic embedding of a fragment is not evident.
-
-\paragraph{Term Rewriting}
-Language-integrated static term rewriting  (``macro'') systems, like Template Haskell \cite{SheardPeytonJones:Haskell-02} and Scala's static macros \cite{ScalaMacros2013}, can be used to decrease complexity when an isomorphic embedding into the underlying type system is possible, but cannot extend the type system directly.% If each metaprogram is invoked like a function, inner macros are expanded before outer macros, and the mechanism enforces abstraction barriers by ensuring that the rewriting logic is hygienic, cannot modify surrounding code, and does not depend on shared state, then compositional reasoning is possible: the meaning of a term depends only on its subterms, not on the specific position it appears in a program.% Complex macros can, however, still be difficult to reason about, so many static macro systems impose further constraints (e.g.
-Scala recommends using only its ``black box'' macros, which  enforce a function-like typing discipline).
-These do not permit extensions to the underlying type system.
-%Rewriting systems that permit global pattern-based dispatch, however, do not admit strong modular reasoning principles (the same term might match multiple patterns, defined separately, creating ambiguities).
-
-% \paragraph{Desugaring}\label{desugaring}
-% Sometimes, dialects address issues of complexity by introducing new syntax.
-Indeed, most dialects do build in syntax for a few privileged abstractions (e.g.
-list literals are nearly ubiquitous, monad comprehensions support a key feature of Haskell).
-%For example, even though lists can be defined using datatypes, most languages build in special ``literal forms'' that make introducing and pattern matching over lists less tedious.
-Different dialects may similarly choose to  provide concise forms for working with other types of data (e.g.
-option types, regular expressions or SQL queries).
-Types not in the language's standard library, however, must typically use a more uniform syntax.
-% %Evidence suggests that this leads programmers to choose less semantically meaningful representations (e.g. representing SQL queries as strings), which can cause a number of issues \cite{TSLs}. 
-% To make the situation less asymmetric, systems that introduce new  ``desugarings'' atop a base language have been developed.
-For example, %Camlp4 is a language-external system used by the OCaml community.
-% dialects of languages built using Sugar* \cite{erdweg2013framework} allow syntax extensions to be packaged separately and combined using language-integrated declarations.
-If the desugaring logic obeys the same constraints described above for macros, semantic reasoning can be performed compositionally (and recent work has shown how reasoning about type correctness can be automated \cite{conf/icfp/LorenzenE13}).
-
-% These systems do not guarantee that the composition of unambiguous grammars will remain unambiguous (and realistic examples of ambiguous combinations are not uncommon, e.g.
-XML and HTML).
-%Every combination of desugarings is thus a  dialect with respect to syntactic reasoning.
-% Recent work has made progress in addressing this problem by restricting how syntax extensions can interface with the host syntax.
-Schwerdfeger and Van Wyk require a globally unique start token and describe checkable conditions pertaining to the follow sets of host language non-terminals to guarantee that extensions can be composed unambiguously \cite{conf/pldi/SchwerdfegerW09}.
-This suffices for modularly adding new keyword-prefixed forms, but literal forms are awkward to define in this way.
-Omar et al.
-describe a language-integrated technique specifically for this purpose, using a technique with parallels to the one we are building on \cite{TSLs}.
-Literal parsing logic is directly associated with user-defined types, forming \emph{type-specific languages} (TSLs), and local type inference controls invocation of TSL parsing logic, guaranteeing that composition is unambiguous.
-The mechanism guarantees hygiene and inner terms cannot be inspected, so  fully modular reasoning is possible.
-
-%We assumed these desugaring techniques were available to solve issues strictly related to complexity and convenience.
-
-\paragraph{Optimization}
-When a simple embedding that preserves the static semantics exists, but a different  embedding would better preserve a desired cost semantics, term rewriting techniques can also be used to perform ``optimizations'', thus achieving an isomorphic embedding.
-Care must be taken, however, to ensure that the optimized value is not manipulated directly if the rewriting does not preserve the static semantics (i.e.
-the rewritten term has a different, less constrained type).
-Type abstraction has been used directly to achieve this property in work on \emph{lightweight modular staging} \cite{Rompf:2012:LMS}.
-This does not allow new type systems to be defined (in LMS, Scala's type system).
-%As long as transformations are meaning-preserving (which can also be shown mechanically), they can be applied in any order.
-%The metaprograms can modularly be shown not to violate the type system of the host language, Modular reasoning is possible as long as dispatch is explicit (e.g.
-macros) or based on an unambiguous language mechanism (e.g.
-trait composition in LMS) and made simpler by the fact that transformations must be type-preserving.
-
-\paragraph{Type Refinements}
-%When T, the solution is less clear. 
-When new static distinctions need to be made for values of an existing type, but new primitive operations are not needed, one solution is to develop a system of \emph{type refinements} based on the fragment of interest \cite{Freeman91}.
-For example, one might refine the type of integers to distinguish negative integers.
-Most proposals for ``pluggable type systems'' describe such type refinement systems \cite{Brac04a}, which require only additional annotations supplied as comments or metadata (e.g.
-JavaCOP \cite{Andreae:2006:FIP:1167473.1167479}).
-Refinement systems do not support holding the refined type abstract as we did in our regular string example ($\tcvar{rstr}$ could have picked a non-string representation for, e.g., fixed size regular languages)%Because the dynamic semantics are fixed,  composition is safe.
-Although our regular string types could perhaps be approximated with refinements, this does not permit the advanced operations, optimizations and representation independence results we enable.
-
-\paragraph{Language Frameworks}
-The most general situation, of which type refinements are a special case, is when exposing a fragment requires defining new types as well as new operators, where the static and dynamic semantics governing these new operators make non-trivial use of  statically valued information.
-We saw labeled product types added, which require a new type constructor and new operator constructors.% The regular string type's group projection operator had a specialized cost semantics, so a simple refinement would not be able to approximate it.
-Scala's type system does not track the necessary invariants, so LMS-like optimizations would also not be appropriate.%For example, projection operators,  written concretely in some dialects of ML as \verb|#label|, can be seen as applications of \verb|#|, the projection operator constructor, to \verb|label|, a static index.
-The type of an operation like \verb|#label e| (sometimes written postfix as \verb|e#label|) is a function of the index of the type of \verb|e|.
-The dynamics also needs (some trace of) this information to determine how the operation is evaluated.
-
-%Many of the other examples mentioned above as motivations for new dialects have a similar flavor.
-We will detail another example in the next section of a type constructor that tracks strings known statically to be in a regular language, using it to dispense with unnecessary run-time checks in cases where they are not strictly necessary (affecting the cost component of the dynamic semantics).
-
-%An important observation is that while it is possible to \emph{implement} records by translation to tuples, or nested binary pairs, or lists, or many other types arising from a language like $\mathcal{L}\{{\rightharpoonup}\,{\forall}\,{\mu}\,{\keyw{1}}\,{\times}\,{+}\}$, it is not possible to isomorphically embed them as such (and thus rely on purely syntactic sugar) because there is no unique inverse mapping from, e.g., unlabeled product types to the corresponding record type (many combinations of labeled products may be implemented using the same unlabeled product).
-This is a common pattern even in more complex situations: a simple internal language (IL) generally suffices as a target for implementing the dynamics of an external language (EL) with a richer static semantics, but an embedding is not possible.
-This observation underlies the design of most compilers.
-
- % like SML, Specifying languages in a manner analagous to how they are usually implemented can simplify reasoning about the core of the language.
-
-% and typed translation semantics (the internal terms have a different type system).% Translation semantics
-
-%These types do not admit an isomorphic embedding using only unlabeled tuples because the field projection family of operators, indexed by labels (field projection is written concretely as \verb|#label(e)|, \verb|e#label| or \verb|e.label| in different languages), must be capable of operating on all values of such types.
-% with the labels removed would not admit a unique inverse.
-%(each equivalence class of record types could be implemented as an abstract type using tuples at considerable inconvenience but there would still be no way to define a projection operator that operated uniformly over all records having a particular field).
-
-% (i.e.
-fragments defining families of operations that cannot simply be written as a finite collection of functions, examples of which we will discuss).
-If there are a finite number of operations, and they can be typechecked without inspecting the structure of the types of the arguments, then an abstract type can be used.
-new primitive operations (including new variants of existing operations) that require static knowledge of these invariants, however, this too is infeasible.
-take advantage of the invariants being maintained, and these operations must be usable throughout a program, .
-%At best, annotations might be written as ``comments'' or metadata for use by these tools.
-
-A variety of \emph{language frameworks} have been developed to make it easier to define new languages, in some cases together with a compiler and other tooling for these languages; see \cite{erdweg2013state}\todo{unlimited references so add a bunch here}.
-%Some \emph{extensible compilers} are also actually language frameworks, because they permit the introduction of new constructs, not just new meaning-preserving optimizations.
-The maintainers of many compilers, e.g.
-Haskell's GHC, treat their compiler as a ``laboratory'' for new fragments, toggled by flags or special comments.
-\emph{Logical frameworks} focus more specifically on helping language designers mechanize the metatheory of language and logic specifications.
-T
-These tools can decrease the effort needed to define a new language, but they either do not support forming languages from separately defined fragments or provide few modular reasoning principles that make it possible to reason separately about these fragments.
-%Every combination of fragments is a new dialect, and must be reasoned about monolithically.
-%Efforts to reason modularly areeed, several problems can come up when fragments are combined, so .
-% Let us briefly review difficulties that arise.
-
-% %Concrete syntax known to be separately unambiguous might not be unambiguous when combined, as was already discussed above. 
-% If the abstract syntax needs to be extended to support a new fragment, problems also arise.
-In monolithic settings, terms can be implemented using finite recursive sums (i.e.
-term constructors are often implemented as ML-style datatype constructors), but this does not permit extension, so open sum types or products of functions (i.e.
-objects \cite{conf/oopsla/Aldrich13}) must instead be used.
-This can present issues when one wishes to modularly define new functionality that should exhaustively cover all terms in the language (e.g.
-pretty-printers for expressions).
-Reynolds first identified this problem \cite{Reynolds75} and Wadler named it the \emph{expression problem}.
-% %A number of language frameworks (e.g.
-JastAdd \cite{Ekman:2007:JEJ:1297027.1297029} and Silver \cite{VanWyk:2010:SEA}) use extensible \emph{attribute grammars}.
-These can be seen as a form of open sum, where attributes correspond to functions performing traversals (more specifically, \emph{catamorphisms} \cite{catamorphisms}\todo{citation / remove?}).
-%When a new term constructor is added, the logic determining how some existing attributes (e.g.
-typechecking and translation) should handle the new case is also provided.
-\emph{Forwarding} can be used to attempt to delegate responsibility over attributes other than those explicitly defined to another term constructor, eventually leading to one in the fixed internal language \cite{VanWyk:2010:SEA}.
-This can address some aspects of the expression problem, but creates a bigger problem: the internal implementation details  of a fragment are necessarily exposed, violating an abstraction barrier that, as we will discuss, is critical.
-% In this work, we sidestep the problems of syntax, instead leaving it fixed and relying on a bidirectional type system delegating to a relevant tycon.
-%For example, projection operators do not require adding a corresponding term constructor to the abstract syntax.
-Instead, projection is categorized as a \emph{targeted operation}, so  the concrete term \verb|#label e| desugars to an abstract term like $\keyw{targ}[(\desugar{\conclbl{\#}}, \desugar{\conclbl{\texttt{label}}})](e)$ (where $\desugar{\cdot}$ denotes an encoding of constant labels, here a label corresponding to the operator constructor and the field label itself, into the static language).
-The type constructor of the type synthesized by the target argument, here $e$, determines the semantics of the term.
-If a new record-like fragment is added (e.g.
-they give the example of one with prototypic inheritance), it can reuse the same concrete and abstract syntax directly, precluding conflicts and avoiding the need to define the behavior of tools like pretty printers for every new fragment.
-As in their work on type-specific languages, described above, the semantics of literal forms are controlled by the type constructor of the type the literal is being analyzed against.
-% %We will discuss this in the next section.
-
-% Once syntactic issues have been addressed, however, there are a host of semantic guarantees  that must be established before a language can be relied upon, as we saw.
-% The most basic guarantee is \emph{type safety}: that the dynamics are well-defined for all terms accepted as well-typed by the statics, and preserve the statics during evaluation.
-Each fragment can then build on type safety to establish additional \emph{type invariants} stating properties about all types that it constructs, which  clients can then use to reason about programs (e.g.
-value induction for eager sum types relies on finiteness of the canonical forms).
-In practice, typechecking is expected be \emph{deterministic} and \emph{decidable} (i.e.
-terminating) and, except in circumstances where non-determinism is explicitly exposed to programmers, the dynamic semantics are also expected to be deterministic.
-%Precise formulations of these properties depend on how the semantics are specified.%, so we will make this more precise below.
-% Modern language frameworks guarantee few or none of these properties about the dialects they produce.
-More alarmingly, even when these properties have been established for two dialects  (either in the metatheory or mechanically using a logical framework), and syntactic conflicts are addressed, there is no guarantee that merging the dialects together will conserve these properties.
-% If our goal is to integrate fragment composition into the language, these tools are thus of limited utility.
-Clients would have to take on the burden of reasoning about these basic properties for every combination of ``libraries'' they chose to import.% Improving this state of affairs, so that it more resembles reasoning about separately defined modules (we assume an ML-style module system), is the general topic of this paper.
-
-% %Though we have not mechanically proven these properties for our design, we have given strong evidence that these properties are maintained modularly.
-That is, we need only establish the semantics of the opcon structures in isolation.
-%We can then rely on, e.g., the Conservativity theorem above.
-
-%We note that is increasingly being deployed in contemporary languages (e.g.
-Scala) due to the perception that it enables good error messages, and to support semantics where  non-local type inference is undecidable.
-Permitting extensibility together with non-local type inference is an open problem.
-\paragraph{Module Systems}
-Our work has some technical similarities to work on maintaining type abstraction using a form of effect system \cite{CraryHarperDreyer2002}, and on translating modules to System F \cite{conf/tldi/RossbergRD10}.
-Unlike modules, tycons are indexed by arbitrary static values, permit the representation of a type to be computed based on this index, and compute both the type and translation of each operation, rather than requiring that the operations be expressible as a finite collection of functions.% This permits us tocode targeting a separate IL, and permit operations whose types cannot simply be written as functions.
-%Both examples we used here would be quite difficult to fully embed using only modules.
-
-\section{Discussion}\label{discussion}
-%The most immediate direction for future work is to finishing mechanizing the metatheory for this  technique, and to embed it into an existing proof assistant (using a continuation passing style to simulate the stores in our normalization semantics).
-Despite this, we believe that the technique as described semi-formally above is compelling.
-Adding support for tycon-specific typing contexts, implicit coercions and direct opcon calls can all be taken up by the community.
-Delving further into the question of when can two tycons with the same signature be substituted for one another, using a technique based on admissible relations, is also an avenue we wish to explore \cite{pfpl}.
-
-%We combine several interesting type theoretic techniques, applying them to novel ends: 1)  a bidirectional type system permits flexible reuse of a fixed syntax; 2) the SL serves both as an extension language and as the type-level language; we give it its own statics (i.e.
-a \emph{kind system}); 3) we use a typed intermediate language and leverage corresponding \emph{typed compilation} techniques, here lifted into the semantics of the EL; 4) we leverage internal type abstraction implicitly as an effect during normalization of the SL to enforce abstraction barriers between type constructors.
-%As a result, conservativity follows from the same elegant parametricity results that underly  abstraction theorems for module systems. 
-%Like modules, reasoning about these \emph{modular type constructors} does not require  mechanized specifications or proofs: correctness issues in the type constructor logic necessarily causes typechecking to fail, so even extensions that are not proven correct can be distributed and ``tested'' in the wild without compromising the integrity of an entire program (at worst, only values of types constructed by the tycon being tested may exhibit undesirable properties).
-
-%\acks
-%The author is grateful to Jonathan Aldrich, Robert Bocchino and anonymous referees for their useful suggestions.
-This work was funded by the DOE Computational Science Graduate Fellowship under grant number DE-FG02-97ER25308.
-%Acknowledgments, if needed.
+\section{External Language}
+\subsection{Additional Desugarings}
+\subsection{Typing}
+\subsection{Proof of Regular String Soundness Tycon Invariant}
 

### 2014-11-02 att-pldi15/att-pldi15.tex
Finished draft of PLDI15 except some citations

@@
 \begin{abstract}
-Researchers often describe typed programming languages as fragments or simple calculi, leaving to language designers the task of composing these to form complete languages.
+Researchers often describe type systems as fragments or simple calculi, leaving to language designers the task of composing these to form complete programming  languages.
 %comparable to those available for library-based embeddings packaged using a modern module system.
 This is not a systematic process: metatheoretic results must be established anew for each composition, guided only notionally by metatheorems derived for simpler systems.
-As the design space grows, mechanisms that provide stronger modular reasoning principles than this are needed.%, so that important metatheoretic results need only be established locally and, ultimately, so that language extensions can be treated like libraries.
+As the language design space grows, mechanisms that provide stronger modular reasoning principles than this are needed.%, so that important metatheoretic results need only be established locally and, ultimately, so that language extensions can be treated like libraries.
 %,  Sometimes this fails because certain language features do not ``play well'' together, but precisely characterizing which is elusive given these practices.
@@
 In this paper, we begin from first principles with a core calculus, @$\lambda$, specified like many full-scale languages: as a bidirectionally typed translation semantics.
-Only the $\rightarrow$ type constructor (tycon) is built in; all other external tycons (we show a variant of record types and constrained string types) are defined by extending a \emph{tycon context}.
-Each tycon defines the semantics of associated term-level operators (e.g.
+Only the $\rightarrow$ type constructor (tycon) is built in; all other external tycons (we show a variant on record types and constrained string types) are defined by extending a \emph{tycon context}.
+Each tycon defines the semantics of its associated term-level operators (e.g.
 record projection) using functions written in a static language where types and translations are values.
 The semantics provide  strong metatheoretic guarantees, notably \emph{type safety} and \emph{conservativity}: that all \emph{tycon  invariants} will be conserved under extension.
-Mechanized proofs are not needed: problems are caught during typechecking by lifting typed compilation techniques into the semantics and enforcing abstraction barriers around tycons using type abstraction, the same principle that underlies ML-style modules.
+Mechanized proofs are not needed: problems are caught during typechecking by lifting typed compilation techniques into the semantics and enforcing barriers around tycons using type abstraction, the same principle that underlies reasoning with ML-style modules.
 \end{abstract}
@@
 
-Universality properties (i.e.
-``Turing-completeness'') can be invoked to guarantee that an embedding that preserves a desirable fragment's dynamic semantics is possible, but an isomorphic embedding must also preserve the static semantics and, if defined, performance bounds specified using a cost semantics.
+Universality properties (e.g.
+``Turing-completeness'') can usually be invoked to guarantee that an embedding that preserves a desirable fragment's dynamics is possible, but an isomorphic embedding must also preserve the static semantics and, if defined, performance bounds specified using a cost semantics.
 This is not always possible.
@@
 %Indeed, the proliferation of language dialects constructed for this reason might be taken as an evidence that the core language is not as ``general'', when considered comprehensively, as might be hoped.
-Each time a language is extended with a new fragment for one of these reasons, a new \emph{dialect} is born. Within the ML lineage, for example, dialects abound:
+Each time a fragment like this is needed, a new \emph{dialect} of a language must be constructed. Within the ML lineage, for example, dialects abound:
 %Reynolds, in a remark that recalls the ``Turing tarpit'' of Perlis \cite{Perl82a}, summarizes the issue \cite{Reynolds94anintroduction}: 
@@
 Using different dialects separately for different components of a program is untenable: components written in different dialects cannot always interface safely (i.e.
-a safe FFI, item 3 above, is needed).
-
-These problems do not merit concern when a fragment is expressed as an isomorphic embedding (i.e.
-as a library) because modern \emph{module systems} enforce abstraction barriers that ensure that the isomorphism need only be established in the ``closed world'' of the module.
+a safe FFI, item 3 above, is needed between every pair of dialects).
+
+These problems do not arise for a fragment that can be expressed as an isomorphic embedding (i.e.
+as a library) because modern \emph{module systems} can enforce abstraction barriers that ensure that the isomorphism need only be established in the ``closed world'' of the module.
 For example, a module defining sets in ML can hold the representation of sets abstract, ensuring that any invariants maintained by the functions in the module (e.g.
@@
 Instead, the approach we take relies on type abstraction in the internal language.
-As a result, we are able to borrow the same results that underly modular reasoning in simply-typed languages like ML to reason modularly about typed language fragments.
-We describe  related work in Sec.
-\ref{prior-work} and conclude with a brief discussion of expressiveness in Sec.
-\ref{discussion}.
+As a result, we are able to borrow the same results that underly modular reasoning in simply-typed languages like ML to reason modularly about language fragments.
+We conclude by discussing  related and future work in Sec.
+\ref{prior-work}.
 
@@
 \hspace{-5px}\input{syntax-table-SL}
-\caption{Syntax of the static language (SL). Metavariable $\sx$ ranges over static term variables, $\kalpha$ and $\kvar$ over kind variables and $m$ and $n$ over natural numbers.}\vspace{-5px}
+\caption{Syntax of the static language (SL). Metavariable $\sx$ ranges over static term variables, $\kalpha$ and $\kvar$ over kind variables and $n$ over natural numbers.}\vspace{-5px}
 \label{syntax-SL}
@@
 \textbf{tycon structures} & \theta & ::= & \tcstruct{\st}{\omega} \\
+\textbf{opcon structures} & \omega & ::= & \tcstructn{\st} ~|~ \tcstructc{\omega}{op}{\st}\\
 \textbf{tycon sigs} & \psi & ::= & \tcsig{\kappa}{\chi}\\
-\textbf{opcon structures} & \omega & ::= & \tcstructn{\st} ~|~ \tcstructc{\omega}{op}{\st}\\
 \textbf{opcon sigs}& \chi & ::= & \introsig{\kappa} ~|~ \opsigS{\theta}{op}{\kappa}\\
@@
 The introductory form for kind $\kty$ is $\sty{c}{\st}$, where $c$ is a \emph{tycon} and $\st$ is the \emph{type index}.
-The syntax  for tycons given in Figure \ref{syntax-TC} specifies that $c$ is either the built-in tycon governing partial functions, $\rightharpoonup$, a user-defined tycon name written in small caps, $\tc$, or an ``other'' tycon, $\keyw{other}[m]$ indexed by a natural number $m$ (to allow arbitrarily many such tycons).
+The syntax  for tycons given in Figure \ref{syntax-TC} specifies that $c$ is either the built-in tycon governing partial functions, $\rightharpoonup$, a user-defined tycon name written in small caps, $\tc$, or an ``other'' tycon, $\keyw{other}[m]$, indexed by a natural number $m$ to allow arbitrarily many such tycons.
 The kinding rules governing the form $\sty{c}{\st}$ are shown in Figure \ref{fig:types}.
 The dynamics are simple (see supplement): the index is eagerly normalized and errors propagate.
-We write $\istype{\st}{\Phi}$  iff $\sofkn{\emptyset}{\emptyset}{\Phi}{0}{\st}{\kty}$ and $\svalNA{\st}$.
+We write $\istype{\st}{\Phi}$  iff $\sofkn{\emptyset}{\emptyset}{\Phi}{0}{\st}{\kty}$ and $\sval{\st}{\cdot;\emptyset;\Phi}$.
 
@@
 \ref{metatheory}.
-The index of such a type must pair a natural number with a type translation of kind $\kity$, discussed in Sec.
+The index of such a type must pair a natural number with a static value of kind $\kity$, discussed in Sec.
 \ref{sec:type-translations}.
@@
 \paragraph{Examples}
-As our first example, consider the user-defined tycon $\tcvar{rstr}$.
-The index kind of $\tcvar{rstr}$ is $\krx$, which classifies static regular expression patterns and we assume is defined as an inductive sum kind in the usual way.
+Our first example of a user-defined tycon is $\tcvar{rstr}$.
+It has index kind $\krx$, which classifies static regular expression patterns (defined as an inductive sum kind in the usual way).
 Types constructed by $\tcvar{rstr}$ will classify \emph{regular strings}, which are statically known to be in the regular language specified by the type index \cite{sanitation-psp14}.
 For example, $\stx{title} := \sty{\tcvar{rstr}}{\concrx{.+}}$ will classify non-empty strings and $\stx{conf} := \sty{\tcvar{rstr}}{\concrx{[A-Z]+ \digit\digit\digit\digit}}$ will classify conference names.
-The type indices are  here written using standard concrete syntax for concision.
-Previous work has shown how to define type-specific (here, kind-specific) syntax like this composably in libraries \cite{TSLs}.
+The type indices are  here written using standard concrete syntax for concision; recent work has specified how to define type-specific (or here, kind-specific) syntax like this composably in libraries \cite{TSLs}.
 We define a tycon context containing only the definition of $\tcvar{rstr}$, $\Phi_\text{rstr} := \tcdef{\tcvar{rstr}}{\tcsig{\krx}{\chi_\text{rstr}}}{\theta_\text{rstr}}$.
@@
 The tycon context containing only $\tcvar{lprod}$'s definition is $\small\Phi_\text{lprod} := \tcdef{\tcvar{lprod}}{\tcsig{\klist{\klbl \times \kty}}{\chi_\text{lprod}}}{\theta_\text{lprod}}$.
-
 In the tycon context containing both tycon definitions, $\Phi_\text{rstr}\Phi_\text{lprod}$, we can define a labeled product type classifying conference papers, $\stx{paper} := \sty{\tcvar{lprod}}{\{\conclbl{title} : \stx{title}, \conclbl{conf} : \stx{conf}\}}$.
@@
 \subsection{Static Type Case Analysis}
-A type $\st$ can be case analyzed using $\stycase{c}{\st}{\sx}{\st_1}{\st_2}$.
+Types in @$\lambda$ can be thought of as arising from a distinguished ``open datatype'' defined by the tycon context \cite{conf/ppdp/LohH06}.
+Consistent with this view, a type $\st$ can be case analyzed using $\stycase{c}{\st}{\sx}{\st_1}{\st_2}$.
 If the value of $\st$ is constructed by $c$, its type index is bound to $\sx$ and the branch $\st_1$ is taken.
@@
 % \end{mathpar}
-The rule for $c{=}{\rightharpoonup}$ is analagous, but no rule for $c{=}\keyw{other}[m]$ is defined (these always take the default branch).
-
-Put another way, types can be thought of as arising from a distinguished ``open datatype'' defined by the tycon context\todo{citation}.
-%The dynamics (see supplement) are again straightforward, consistent with this intuition.% We will see an example of its use in Sec.
-\ref{sec:targops}.
+The rule for $c{=}{\rightharpoonup}$ is analagous, but no rule for $c{=}\keyw{other}[m]$ is defined (such types always take the default branch).
+
+ %The dynamics (see supplement) are again straightforward, consistent with this intuition.% We will see an example of its use in Sec. \ref{sec:targops}.
 \begin{figure}\hfill \fbox{$\vdash \Phi$}\vspace{-25px}\begin{mathpar}
@@
 \subsection{Type Translations}\label{sec:type-translations}
+Recall that every type $\st$ must have a type translation, $\tau$.
 Each tycon in the tycon context {computes} translations for the types it constructs as a function of each type's index by specifying a \emph{translation schema} in the {tycon structure}, $\theta$.
@@
 
-The kinding rules (and dynamics) for these shared forms simply proceed recursively, e.g.
+The kinding and dynamics for these shared forms simply proceed recursively, e.g.
 \begin{mathpar}
@@
 
-The syntax for translational internal types additionally includes an ``unquote'' form,  $\qtuq{\st}$, so that they can be constructed compositionally, as well as a form, $\srep{\st}$, that allows one type's translation to refer to the translation of another type $\st$: \begin{mathpar}\small
+The syntax for translational internal types additionally includes an ``unquote'' form,  $\qtuq{\st}$, so that they can be constructed compositionally, as well as a form, $\srep{\st}$, that allows one type's translation to refer to another: \begin{mathpar}\small
 \inferrule[k-ity-unquote]{
@@
 we could also have used a list).
-We assume $\small\svar{listrec} :: \kforall{\kalpha_1}{\kforall{\kalpha_2}{\klist{\kalpha_1}\rightarrow\kalpha_2\rightarrow(\kalpha_1\rightarrow\kalpha_2\rightarrow\kalpha_2)\rightarrow\kalpha_2}}$ in defining $\theta_\text{lprod} := \keyw{trans}=\stx{lprod/trans}, \omega_\text{lprod}$  where $\stx{lprod/trans}:=$
+We assume $\small\svar{listrec} :: \kforall{\kalpha_1}{\kforall{\kalpha_2}{\klist{\kalpha_1}\rightarrow\kalpha_2\rightarrow(\kalpha_1\rightarrow\kalpha_2\rightarrow\kalpha_2)\rightarrow\kalpha_2}}$ in defining $\theta_\text{lprod} := \keyw{trans}=\stx{lprod/trans}~\keyw{in}~\omega_\text{lprod}$  where $\stx{lprod/trans}:=$
 \[\small
@@
 This can be thought of as analagous to the process in ML by which the true identity of an abstract type in a module is held abstract outside the module until after typechecking.
-The judgement $\small\tdeabs{\Phi}{c}{\qity}{\memD}{\ity\moutput}{\memD\moutput}$ relates a normalized translational internal type $\qity$ to an internal type $\tau$, called a corresponding \emph{abstracted type translation} because references to translations of types constructed by a tycon other than those constructed by a ``delegated tycon'', $c$, are replaced by universally quantified type variables, $\alpha$.
+The judgement $\small\tdeabs{\Phi}{c}{\qity}{\memD}{\ity\moutput}{\memD\moutput}$ relates a normalized translational internal type $\qity$ to an internal type $\tau$, called a \emph{selectively abstracted type translation} because references to translations of types constructed by a tycon other than the ``delegated tycon'', $c$, are replaced by a corresponding type variables, $\alpha$.
+The \emph{type translation store} $\memD ::= \emptyset ~|~ \memD, \st \leftrightarrow \ity/\alpha$ maintains the corresponding between types, their actual translations and the unique type variables which appear in their place.
 For example, $\small\tdeabs{\Phi_\text{rstr}\Phi_\text{lprod}}{\tcvar{lprod}}{\qity_\text{paper/trans}}{\emptyset}{\tau_\text{paper/abs}}{\memD_\text{paper/abs}}$
@@
 $\memD_\text{paper/abs}  := \stx{title} \leftrightarrow \keyw{str}/\alpha_1, \stx{conf} \leftrightarrow \keyw{str}/\alpha_2$.
-The \emph{type translation store} $\memD ::= \emptyset ~|~ \memD, \st \leftrightarrow \ity/\alpha$ maintains the  mapping from types, $\st$, to their (non-abstract) translations, $\tau$, and the type variables, $\alpha$, which appear in place of their translation.
 
@@
 For example, $[\delta_\text{paper/abs}]\tau_\text{paper/abs}$ is $\tau_\text{paper} := \keyw{str} \times (\keyw{str} \times \iunit)$, i.e.
-the final type translation of $\stx{paper}$.
+the actual type translation of $\stx{paper}$.
 Indeed, we can now give the rule defining the type translation judgement, $\vdash_\Phi \st \leadsto \tau$, mentioned in Sec.
@@
 
-With this intuition, we can now give the rules for the selective type abstraction judgement.
-It first recurses generically over sub-terms of $\qity$ until sub-terms of form $\srep{\st}$ are encountered (see supplement).
+The rules for the selective type abstraction judgement are straightforward.
+We first recurse generically over sub-terms of $\qity$ until sub-terms of form $\srep{\st}$ are encountered (see supplement).
 The translation of partial function types is direct and is not held abstract, so that lambdas can be used as the sole binding construct in the EL:
@@
 \subsection{Typing Contexts}
-We must also define a translation judgement for external typing contexts $\vdash_\Phi \Upsilon \leadsto \Gamma$ that checks that $\Upsilon$ contains valid types, then generates the corresponding internal typing context $\Gamma$ from their translations (see supplement).
+External typing contexts contain types, so we can now also define a translation judgement  $\vdash_\Phi \Upsilon \leadsto \Gamma$ (see supplement).
 %\vspace{-5px}
@@
 
-Having established how types are constructed, and how they determine abstract and from there concrete translations to internal types, we can finally give the typing and translation rules for external terms, shown in Figure \ref{typing}.
+Having established how types are constructed, and how they determine selectively abstract and from there actual type translations, we can finally give the typing and translation rules for external terms, shown in Figure \ref{typing}.
 
@@
 Before discussing rules (ana-intro) and (ana-intro-other), we note that we can recover a variety of standard concrete introductory forms by a purely syntactic desugaring to this abstract form (and thus allow their use at more than one type).
-For example, for regular strings we can use the string literal form, $\texttt{"s"}$, which desugars to $\eintro{\texttt{"s"}_\text{SL}}{\cdot}$.
-The term index is the corresponding static value of kind $\keyw{Str}$, indicated by a subscript for clarity.
+For example, for regular strings we can use the string literal form, $\texttt{"s"}$, which desugars to $\eintro{\texttt{"s"}_\text{SL}}{\cdot}$, i.e.
+the term index is the corresponding static value of kind $\keyw{Str}$, indicated by a subscript for clarity.
 Similarly, for labeled products, records, objects and so on, we can define a generalized labeled collection form, $\{\mathtt{lbl}_1=e_1, \ldots, \mathtt{lbl}_n=e_n\}$, that desugars to $\eintro{\texttt{[}\mathtt{lbl}_1, \ldots, \mathtt{lbl}_n\texttt{]}}{e_1; \ldots; e_n}$, i.e.
@@
 %The supplement gives the analagous rules.
-
 The final two forms of translational internal term are $\anatrans{n}{\st}$ and $\syntrans{n}$.
@@
 Then, we request type analysis of the corresponding argument, $\svar{rowarg}$,  against the type in the type index, $\svar{rowty}$, by writing $\svar{ana}~\svar{rowarg}~\svar{rowty}$.
-Here, $\svar{ana}$ is a helper function defined below that produces a translational internal term  of the form $\sqitm{\anatrans{n}{\st}}$, where $\st$ is the value of $\svar{rowty}$, if the row value is well-typed, and raises an error if not.
+Here, $\svar{ana}$ is a helper function defined below that produces a translational internal term  of the form $\sqitm{\anatrans{n}{\st}}$, where $\st$ is the value of $\svar{rowty}$, if the argument is well-typed, and raises an error if not.
 The final line constructs a nested tuple based on this translation and the recursive result.
-Taken together, the translational internal term that will be derived for our example involving $e_\text{ex}$ above is $\small\qitm_\text{ex} := (\anatrans{0}{\stx{title}}, (\anatrans{1}{\stx{conf}}, ()))$.
+Taken together, the translational internal term that will be generated for our example involving $e_\text{ex}$ above is $\small\qitm_\text{ex} := (\anatrans{0}{\stx{title}}, (\anatrans{1}{\stx{conf}}, ()))$.
 
@@
 
-The argument interface list, called $\stx{args}$ in the rules, is simply a static list of kind $\kargs$ where the $i$th entry is $(\slam{\kty}{\svar{ty}}{\sana{i}{\svar{ty}}}, \slam{\kunit}{\_}{\ssyn{i}})$.
-It is generated by the judgement $\keyw{args}(n)=\stx{args}$, where $n$ is given as the length of the argument list, written $|\es|=n$ (see supplement).
+The argument interface list generated by the judgement $\keyw{args}(n)=\stx{args}$, where $n$ is the length of the argument list, written $|\es|=n$, is then simply a static value of kind $\kargs$ of length $n$ where the $i$th entry is $(\slam{\kty}{\svar{ty}}{\sana{i}{\svar{ty}}}, \slam{\kunit}{\_}{\ssyn{i}})$.
 
@@
 
-The dynamics for these forms couples the dynamics of the static language to the statics of the external language.
+These forms serve as the link between the dynamics of the static language and the statics of the external language.
 For $\sana{n}{\st}$, after stepping the type $\st$ to a value and propagating errors, the argument environment, which stores the arguments themselves and the typing and tycon contexts, $\es; \Upsilon; \Phi$, is consulted to retrieve the $n$th argument and analyze it against $\st$.
@@
 \inferrule[s-ana-success]{
-    \svalNA{\st}\\
+    \sval{\st}{\es;\Upsilon;\Phi}\\
     \keyw{nth}[n](\es) = e\\
@@
 \inferrule[s-ana-fail]{
-    \svalNA{\st}\\
+    \sval{\st}{\es;\Upsilon;\Phi}\\
     \keyw{nth}[n](\es) = e\\
@@
 
-The kinding rules also prevent the translational internal term forms generated by these static operations from being well-kinded when $n = 0$.
+The kinding rules also prevent these translational internal term forms  from being well-kinded when $n = 0$.
 Like the translational internal type form $\srep{\st}$, these  are retained in values of kind $\kitm$:
@@
 
-\paragraph{Abstracted Term Translations} The reason for this  is again because we will hold these abstract by replacing them with variables.
-The judgement $\edeabs{\tc}{\argEnv}{\qitm}{\memD}{\memG}{\itm\moutput}{\memD'\moutput}{\memG'\moutput}$, appearing as the eighth premise of (ana-intro), relates a translational internal term $\qitm$ to an internal term $\itm$ called the corresponding \emph{abstracted term translation}, where all references to the translation of an argument (having any type) are replaced with a unique variable and all references to the translation of a type constructed by a user-defined tycon other than the ``delegated tycon'' $\tc$ are replaced with an abstract type variable as described in Sec.
+\paragraph{Selectively Abstracted Term Translations} The reason for this  is again because we will hold these abstract by replacing them with variables.
+The judgement $\edeabs{c}{\argEnv}{\qitm}{\memD}{\memG}{\itm\moutput}{\memD\moutput}{\memG\moutput}$, appearing as the eighth premise of (ana-intro), relates a translational internal term $\qitm$ to an internal term $\itm$ called its \emph{selectively abstracted term translation}, where all references to the translation of an argument (having any type) are replaced with a unique variable and all references to the translation of a type constructed by a user-defined tycon other than the ``delegated tycon'' $\tc$ are replaced with an abstract type variable as described in Sec.
 \ref{sec:selective-type-translation-abstraction}.
@@
 In other words, the translation of the labeled product $e_\text{ex}$ generated by $\tcvar{lprod}$ is checked with all references to term and type translations of regular strings replaced by variables and type variables, respectively.
-Nevertheless, because our intro opcon definition treated arguments parametrically, the check succeeds.
-We will describe  an ill-behaved operator, which would lead to a violation of the regular string tycon  invariant were it not disallowed by this check,  in Sec.
+Because our intro opcon definition treated arguments parametrically, the check succeeds.
+We  describe  an ill-behaved operator in Sec.
 \ref{metatheory}.
 
-Applying the substitutions $\gamma_\text{ex/abs}$ and $\delta_\text{ex/abs}$ in the conclusion of the rule, we arrive at $\iota_\text{ex}$.
+Applying the substitutions $\gamma_\text{ex/abs}$ and $\delta_\text{ex/abs}$ in the conclusion of the rule, we arrive at the actual term translation $\iota_\text{ex}$.
 
@@
 The remaining premises of (syn-targ) are analagous to the corresponding premises in (ana-intro), with the only difference being that we check the abstract term translation against the abstract type translation of the synthesized type.
-Like (ana-intro-other), the rule (syn-targ-other) is used when the target synthesizes an ``other'' type.
-The mapping from the arguments to a type and translation is given directly in the term index (the operator name is ignored).
+
+\paragraph{Other Targeted Ops} Like (ana-intro-other), rule (syn-targ-other) is used when the target synthesizes an ``other'' type.
+The mapping from the arguments to a type and translation is given directly in the term index (the op name is ignored).
 
@@
 \paragraph{Kind Safety} Kind safety ensures that normalization of well-kinded static terms cannot go wrong. We can take a standard progress and preservation based approach. 
-\begin{theorem}[Static Progress]
-If $\sofkn{\emptyset}{\emptyset}{\Phi}{n}{\st}{\kappa}$ and $\vdash \Phi$ and $|\es|=n$ and $\vdash_\Phi \Upsilon \leadsto \Gamma$ then $\sval{\st}{\es; \Phi; \Upsilon}$ or $\serr{\st}{\es; \Phi; \Upsilon}$ or $\sstep{\st}{\es; \Phi; \Upsilon}{\st'}$.
+\begin{theorem}[Static Progress]\label{thm:static-progress}
+If $\sofkn{\emptyset}{\emptyset}{\Phi}{n}{\st}{\kappa}$ and $\vdash \Phi$ and $|\es|=n$ and $\vdash_\Phi \Upsilon \leadsto \Gamma$ then $\sval{\st}{\es; \Upsilon; \Phi}$ or $\serr{\st}{\es; \Upsilon; \Phi}$ or $\sstep{\st}{\es; \Upsilon; \Phi}{\st'}$.
 \end{theorem}
 
-\begin{theorem}[Static Preservation]
-If $\sofkn{\emptyset}{\emptyset}{\Phi}{n}{\st}{\kappa}$ and $\vdash \Phi$ and $|\es|=n$ and $\vdash_\Phi \Upsilon \leadsto \Gamma$ and $\sstep{\st}{\es; \Phi; \Upsilon}{\st'}$ then $\sofkn{\emptyset}{\emptyset}{\Phi}{n}{\st'}{\kappa}$.
+\begin{theorem}[Static Preservation]\label{thm:static-preservation}
+If $\sofkn{\emptyset}{\emptyset}{\Phi}{n}{\st}{\kappa}$ and $\vdash \Phi$ and $|\es|=n$ and $\vdash_\Phi \Upsilon \leadsto \Gamma$ and $\sstep{\st}{\es; \Upsilon; \Phi}{\st'}$ then $\sofkn{\emptyset}{\emptyset}{\Phi}{n}{\st'}{\kappa}$.
 \end{theorem}
-The case in the proof for the form $\ssyn{n}$ requires that the following theorem be simultaneously defined.
+The case in the proof of Theorem \ref{thm:static-preservation} for $\ssyn{n}$ requires that the following theorem be simultaneously defined.
 The mutual induction is well-founded because the total number of argument lists in the terms being considered decreases.
@@
 Type safety for the IL \cite{pfpl} then implies that well-typed external terms cannot go wrong.
-To prove this, we must prove a stronger theorem, \emph{type-preserving translation}, which is the analog of type-preserving compilation in the typed compilation literature \cite{tarditi+:til-OLD}\todo{right citation for TIL?}:% produced as the translation of the external type.
+To prove this, we must actually prove a stronger theorem, \emph{type-preserving translation}, analagous to type-preserving compilation in the typed compilation literature \cite{tarditi+:til-OLD}\todo{right citation for TIL?}:% produced as the translation of the external type.
 Note that we need only to state our top-level theorems for the analysis judgement  because of the subsumption rule.
@@
 The interesting cases are (ana-intro), (ana-intro-other), (syn-trans) and (syn-trans-other); the latter two arise via subsumption.
-The result follows directly from the final premise of each rule, combined with lemmas that states that if $\memD \leadsto \delta : \Delta_\text{abs}$ and $\memG \leadsto \gamma : \Gamma_\text{abs}$ then $\vdash \delta : \Delta$ and $\Delta~\Gamma \vdash \gamma : \Gamma_\text{abs}$, i.e.
-that all the variables in the contexts have well-formed/well-typed substitutions, and thus that applying them in the conclusions of the rules gives a well-typed term.
+The result follows directly from the final premise of each rule, combined with lemmas that state that if $\memD \leadsto \delta : \Delta_\text{abs}$ and $\memG \leadsto \gamma : \Gamma_\text{abs}$ then $\vdash \delta : \Delta$ and $\Delta_\text{abs}~\Gamma \vdash \gamma : \Gamma_\text{abs}$, i.e.
+that all the variables in $\Delta_\text{abs}$ and $\Gamma_\text{abs}$ have well-formed/well-typed substitutions in $\delta$ and $\gamma$,  and thus that applying them in the conclusions of the rules gives a well-typed term.
 \end{proof-sketch}
@@
 
-\paragraph{Unicity}
-The rules are structured so that if a term is well-typed, both its type and translation are unique.\todo{could move this whole thing to supplement if room needed}
-\begin{theorem}[Unicity]
-If $\vdash \Phi$ and $\vdash_\Phi \Upsilon \leadsto \Gamma$ and $\vdash_\Phi \st \leadsto \tau$ and $\vdash_\Phi \st' \leadsto \tau'$ and $\eanaX{e}{\st}{\iota}$ and $\eanaX{e}{\st'}{\iota'}$ then $\st = \st'$ and $\tau=\tau'$ and $\iota = \iota'$.
-\end{theorem}
-
 \paragraph{Stability}
@@
 \begin{theorem}[Stability]
-Letting $\Phi' := \Phi, \tcdef{\tc}{\psi}{\theta}$, if $\vdash \Phi'$ and $\vdash_\Phi \Upsilon \leadsto \Gamma$ and $\vdash_\Phi \st \leadsto \tau$ and $\eanaX{e}{\st}{\iota}$ then, $\vdash_{\Phi'} \Upsilon \leadsto \Gamma$ and $\vdash_{\Phi'} \st \leadsto \tau$ and $\eana{\Upsilon}{\Phi'}{e}{\st}{\iota}$.
+Letting $\Phi' := \Phi, \tcdef{\tc}{\psi}{\theta}$, if $\vdash \Phi'$ and $\vdash_\Phi \Upsilon \leadsto \Gamma$ and $\vdash_\Phi \st \leadsto \tau$ and $\eanaX{e}{\st}{\iota}$ then $\vdash_{\Phi'} \Upsilon \leadsto \Gamma$ and $\vdash_{\Phi'} \st \leadsto \tau$ and $\eana{\Upsilon}{\Phi'}{e}{\st}{\iota}$.
 \end{theorem}
@@
 Such a calculus and compiler specification was given by Fulton et al.
-\cite{sanitation-psp14} and extending this to our  setting requires showing only that the opcon definitions in $\tcvar{rstr}$ adequately capture these specifications using standard program correctness techniques for the SL, which is a simple functional language.
+\cite{sanitation-psp14} and extending this to our  setting requires showing only that the opcon definitions in $\tcvar{rstr}$ adequately capture these specifications using standard program correctness techniques for the SL, which is a simply-typed functional language.
 The only ``wrinkle'' is that the rule (syn-targ-other) can synthesize a regular string type.
-Luckily the abstracted translation will always be checked against $\tau_\text{abs}=\alpha$, so no strings that could not arise by another rule could  be generated, by the standard abstraction theorem for the IL.
-\end{proof-sketch}
+But if so, the abstracted translation will be checked against $\tau_\text{abs}=\alpha$, so only strings that could otherwise arise can result, a  consequence of the standard abstraction theorem for the IL.\todo{citation} \end{proof-sketch}
 
@@
 $\stx{paper}$, and generating a translation that is not in the corresponding regular language, e.g.
-$\texttt{""}$, could be defined, but when used, the rule (syn-targ) would check the translation against $\tau_\text{abs}=\alpha$, so the invariant could not be broken by the abstraction theorem for the IL.
+$\texttt{""}$, could be defined, but when used, the rule (syn-targ) would check the translation against $\tau_\text{abs}=\alpha$, which would fail.
 %Type abstraction is the basis for conservatively composing type system fragments, just like it is the basis for composing ML-style modules.
- We can state this conservativity property more generally:
+ We can state this more generally:
 
 \begin{theorem}[Conservativity] If $\vdash \Phi$ and $\tc \in \text{dom}(\Phi)$ and a tycon invariant for $\tc$ holds under $\Phi$: \begin{itemize}
-\item If $\eana{\emptyset}{\Phi}{e}{\sty{\tc}{\sttyidx}}{\iota}$ and $\vdash_\Phi {\sty{\tc}{\sttyidx}} \leadsto \tau$ and $\iota \Downarrow \iota'$ then $P(\sttyidx, \iota')$.
+\item If $\eana{\emptyset}{\Phi}{e}{\sty{\tc}{\sttyidx}}{\iota}$ and $\vdash_\Phi {\sty{\tc}{\sttyidx}} \leadsto \tau$  then $P(\sttyidx, \iota)$.
 \end{itemize} then for all $\Phi' = \Phi, \tcdef{\tc'}{\psi'}{\theta'}$ such that $\vdash \Phi'$, the same tycon invariant holds under $\Phi'$: \begin{itemize}
-\item If $\eana{\emptyset}{\Phi'}{e}{\sty{\tc}{\sttyidx}}{\iota}$ and $\vdash_{\Phi'}{\sty{\tc}{\sttyidx}}\leadsto{\tau}$ and $\iota \Downarrow \iota'$ then $P(\sttyidx, \iota')$.
+\item If $\eana{\emptyset}{\Phi'}{e}{\sty{\tc}{\sttyidx}}{\iota}$ and $\vdash_{\Phi'}{\sty{\tc}{\sttyidx}}\leadsto{\tau}$ then $P(\sttyidx, \iota)$.
 \end{itemize}
+(if proposition $P(\sttyidx, \iota)$ is \emph{modular}, defined below)
 \end{theorem}
 \begin{proof-sketch}
-The proof maps every well-typed term in the extended tycon context to a well-typed term in the original tycon context with the same translation, so we can apply the original invariant to arrive at the property $P(\sttyidx, \iota')$.
-This term is constructed by replacing all types constructed by $\tc'$ with a type constructed by $\keyw{other}[m]$, for an $m$ uniquely associated with $\tc'$, and all introductory and targeted sub-terms associated with $\tc'$ with an equivalent one that passes through the rules (ana-intro-other) and (syn-targ-other) by simply partially applying the intro and targeted opcon definitions to generate the term indices.
-Types constructed by $\keyw{other}[m]$ are indistinguishable from types constructed by a ``future'' tycon because the type case analysis construct is not defined for this tycon form, so the translations coincide.\end{proof-sketch}
+The proof maps every well-typed term under $\Phi'$ to a well-typed term under $\Phi$ with the same translation, and if the term has a type constructed by a tycon in $\Phi$, e.g.
+$\tc$, the mapping still has a type constructed by that tycon with the same type translation and thus we can apply the original tycon invariant.
+The only condition on $P(\sttyidx, \iota)$ is that it be preserved under this mapping on $\sttyidx$.
+We call such propositions \emph{modular}.
+Non-modular propositions are  uninteresting because the mapping's only effect on static terms is to replace all types constructed by $\tc'$ with a unique type constructed by $\keyw{other}[m]$, for an $m$ uniquely associated with $\tc'$.
+On external terms, the mapping replaces all introductory and targeted sub-terms associated with $\tc'$ with an equivalent one that passes through the rules (ana-intro-other) and (syn-targ-other) by partially applying the intro and targeted opcon definitions to generate the term indices.
+Typing, kinding, static normalization and the selective translation abstraction judgements are all preserved under this mapping, defined inductively in the supplement.
+\end{proof-sketch}
 %Had the translational term generated by the intro opcon definition been, e.g., $(\anatrans{0}{\stx{title}}, (\texttt{"TEST"}, ())$, then the corresponding abstract term would be $(x_0, (\texttt{"TEST"}, ())$ and the check would fail because $(\alpha_0, \alpha_1)~(x_0 : \alpha_0) \nvdash \texttt{"TEST"} : \alpha_1$.
@@
 
-\section{Related Work}\label{prior-work}
-
-% Note also that type constructors are not themselves static values.
-To apply them in a curried manner or pass them around in the static language, we can define a static function, e.g.
-$\slam{\kty}{\svar{t1}}{\slam{\kty}{\svar{t2}}{\sparr{(\svar{t1}, \svar{t2})}}}$ has kind $\karrow{\kty}{\karrow{\kty}{\kty}}$.
-
-%To situate our work, and because we build on some recent results, we begin by summarizing prior approaches that can be considered when a simple isomorphic embedding of a fragment is not evident.
-
-\paragraph{Term Rewriting}
+\section{Related and Future Work}\label{prior-work}
+As discussed throughout the paper, our treatment of concrete syntax defers to recent work on \emph{type-specific languages} \cite{TSLs}.
+As in our work, TSLs determine the meaning of introductory terms based on the type the term is analyzed against, but TSLs only support syntactic elaboration to a fixed semantics.
+
+%\paragraph{Term Rewriting}
 Language-integrated static term rewriting  (``macro'') systems, like Template Haskell \cite{SheardPeytonJones:Haskell-02} and Scala's static macros \cite{ScalaMacros2013}, can be used to decrease complexity when an isomorphic embedding into the underlying type system is possible, but cannot extend the type system directly.% If each metaprogram is invoked like a function, inner macros are expanded before outer macros, and the mechanism enforces abstraction barriers by ensuring that the rewriting logic is hygienic, cannot modify surrounding code, and does not depend on shared state, then compositional reasoning is possible: the meaning of a term depends only on its subterms, not on the specific position it appears in a program.% Complex macros can, however, still be difficult to reason about, so many static macro systems impose further constraints (e.g.
@@
 
-\paragraph{Optimization}
-When a simple embedding that preserves the static semantics exists, but a different  embedding would better preserve a desired cost semantics, term rewriting techniques can also be used to perform ``optimizations'', thus achieving an isomorphic embedding.
-Care must be taken, however, to ensure that the optimized value is not manipulated directly if the rewriting does not preserve the static semantics (i.e.
-the rewritten term has a different, less constrained type).
-Type abstraction has been used directly to achieve this property in work on \emph{lightweight modular staging} \cite{Rompf:2012:LMS}.
-This does not allow new type systems to be defined (in LMS, Scala's type system).
+%\paragraph{Optimization}
+When a simple embedding that preserves the static semantics exists, but a different  embedding is needed to preserve a desired cost semantics, term rewriting  can also be used to perform ``optimizations'', thus achieving an isomorphic embedding.
+Care is needed when the optimized value has a different type, to ensure that the invariants of the unoptimized term are preserved.
+Type abstraction has been used for this purpose in work on \emph{lightweight modular staging} (LMS) \cite{Rompf:2012:LMS}.
+This does not enable new type systems, i.e.
+in \cite{Rompf:2012:LMS}, Scala's type system must suffice.
 %As long as transformations are meaning-preserving (which can also be shown mechanically), they can be applied in any order.
@@
 
-\paragraph{Type Refinements}
+%\paragraph{Refinement Types}
 %When T, the solution is less clear. 
-When new static distinctions need to be made for values of an existing type, but new primitive operations are not needed, one solution is to develop a system of \emph{type refinements} based on the fragment of interest \cite{Freeman91}.
-For example, one might refine the type of integers to distinguish negative integers.
-Most proposals for ``pluggable type systems'' describe such type refinement systems \cite{Brac04a}, which require only additional annotations supplied as comments or metadata (e.g.
-JavaCOP \cite{Andreae:2006:FIP:1167473.1167479}).
-Refinement systems do not support holding the refined type abstract as we did in our regular string example ($\tcvar{rstr}$ could have picked a non-string representation for, e.g., fixed size regular languages)%Because the dynamic semantics are fixed,  composition is safe.
+When new static distinctions are needed within an existing type, but new operators are not necessary and cost is acceptable, one solution is to develop a system of \emph{refinement types} for the language \cite{Freeman91}.
+For example, a refinement of  integers might distinguish negative integers.
+Proposals for ``pluggable type systems'' describe such type refinement systems \cite{Brac04a,Andreae:2006:FIP:1167473.1167479}.
+Refinements of abstract types can be used to hold representations abstract.
+%Because the dynamic semantics are fixed,  composition is safe.
 Although our regular string types could perhaps be approximated with refinements, this does not permit the advanced operations, optimizations and representation independence results we enable.
 
-\paragraph{Language Frameworks}
-The most general situation, of which type refinements are a special case, is when exposing a fragment requires defining new types as well as new operators, where the static and dynamic semantics governing these new operators make non-trivial use of  statically valued information.
-We saw labeled product types added, which require a new type constructor and new operator constructors.% The regular string type's group projection operator had a specialized cost semantics, so a simple refinement would not be able to approximate it.
+%\paragraph{Language Frameworks}
+Our interest is in a more general situation, of which refinement types can be seen in ways as a degenerate case: when introducing a fragment requires defining new types and new operators, where the types and translations arising from the new operators are a  function of  statically valued information.
+For example, labeled products require new opcons that use labels and types, which exist only statically.
+%Regular string types could also support an operation that the internal string type would not (e.g.
+extraction of a subgroup defined in the type index).% The regular string type's group projection operator had a specialized cost semantics, so a simple refinement would not be able to approximate it.
 Scala's type system does not track the necessary invariants, so LMS-like optimizations would also not be appropriate.%For example, projection operators,  written concretely in some dialects of ML as \verb|#label|, can be seen as applications of \verb|#|, the projection operator constructor, to \verb|label|, a static index.
@@
 
-A variety of \emph{language frameworks} have been developed to make it easier to define new languages, in some cases together with a compiler and other tooling for these languages; see \cite{erdweg2013state}\todo{unlimited references so add a bunch here}.
+Many \emph{language frameworks} and \emph{compiler generators} exist to simplify  dialect implementation (e.g.
+\todo{citations}).
 %Some \emph{extensible compilers} are also actually language frameworks, because they permit the introduction of new constructs, not just new meaning-preserving optimizations.
@@
 T
-These tools can decrease the effort needed to define a new language, but they either do not support forming languages from separately defined fragments or provide few modular reasoning principles that make it possible to reason separately about these fragments.
-%Every combination of fragments is a new dialect, and must be reasoned about monolithically.
+These either do not support forming languages from separately defined fragments or provide few clear modular reasoning principles that make it possible to reason separately about them.
+
+Proof assistants and logical frameworks can be used to mechanize dialect specification and metatheory, but they similarly require a complete specification.
+Chlipala has suggested using  proof automation that covers a variety of language features to decrease the burden when a  dialect is formed, but this is a fundamentally heuristic approach \cite{Chlipala10}\todo{wrong citation}.
+Our approach provides a conservativity theorem.
+No proof terms are needed to guarantee modular reasoning:   mistakes can only weaken the invariants at a single tycon.
+%This facilitates exploratory language design and suggests educational applications.
+A promising avenue for future work beyond the scope of this paper is to introduce optional proof mechanization into the SL (by making it dependently kinded), and to mechanize the metatheory of @$\lambda$ itself.%Every combination of fragments is a new dialect, and must be reasoned about monolithically.
 %Efforts to reason modularly areeed, several problems can come up when fragments are combined, so .
@@
 Permitting extensibility together with non-local type inference is an open problem.
-\paragraph{Module Systems}
-Our work has some technical similarities to work on maintaining type abstraction using a form of effect system \cite{CraryHarperDreyer2002}, and on translating modules to System F \cite{conf/tldi/RossbergRD10}.
-Unlike modules, tycons are indexed by arbitrary static values, permit the representation of a type to be computed based on this index, and compute both the type and translation of each operation, rather than requiring that the operations be expressible as a finite collection of functions.% This permits us tocode targeting a separate IL, and permit operations whose types cannot simply be written as functions.
+
+Type abstraction is, encouragingly, also the technique underlying modular reasoning in ML-like languages.
+Indeed, proofs of tycon invariants rely on the same abstraction theorem.
+Our work has some technical similarities to work on syntactic type abstraction \cite{journals/toplas/GrossmanMZ00} and on translating modules to System F \cite{conf/tldi/RossbergRD10}.
+Unlike in module systems, type translations (analagous to the choice of representation for an abstract type) can be \emph{computed} based on a type index.
+Moreover, there can be arbitrarily many operators because they arise by providing a term index to an opcon.
+Their static and dynamic semantics can be complex because a static function computes the type and translation.
+In contrast, modules can only specify a fixed number of operations, and each must have function type.
+Note that although we did not specify quantification over external types here for simplicity, we conjecture that it is complementary (and thus @$\lambda$ could serve as the core of a language with a ML-style module system).
+Another avenue for future work is tycon functors, which abstract over tycons with the same tycon signature.
+% This permits us tocode targeting a separate IL, and permit operations whose types cannot simply be written as functions.
 %Both examples we used here would be quite difficult to fully embed using only modules.
 
-\section{Discussion}\label{discussion}
+One limitation of our approach is that it supports only  fragments with the same ``shape'' of typing judgement.
+Fragments that require new forms of contexts (e.g.
+symbol contexts \cite{pfpl})  cannot be defined.
+The language controls variable binding, so new notions of binding, e.g.
+linear type systems, cannot be defined.
+Another limitation is that opcons cannot invoke one another (e.g.
+a ``length'' opcon on regular strings could not construct a natural number).
+We conjecture that these are not fundamental limitations.
+%(tycons could define new contexts which are threaded opaquely through ``outside'' operations).
+
 %The most immediate direction for future work is to finishing mechanizing the metatheory for this  technique, and to embed it into an existing proof assistant (using a continuation passing style to simulate the stores in our normalization semantics).
@@
 
-\newpage
-\appendix
-\section{Appendix}
-\begin{mathpar}
-\small
-\inferrule[s-ty-step]{
-    \sstep{\st}{\argEnv}{\st'}
-}{
-    \sstep{\sty{c}{\st}}{\argEnv}{\sty{c}{\st'}}
-}
-
-\inferrule[s-ty-err]{
-    \serr{\st}{\argEnv}
-}{
-    \serr{\sty{c}{\st}}{\argEnv}
-}
-
-\inferrule[s-ty-v]{
-    \sval{\st}{\argEnv}
-}{
-    \sval{\sty{c}{\st}}{\argEnv}
-}
-
-\inferrule[s-otherty-v]{ }{
-    \sval{\sotherty{m}{\tau}}{\argEnv}
-}
-\end{mathpar}
-
-\begin{mathpar}
-\small
-\inferrule[s-tycase-step]{
-    \sstep{\st}{\argEnv}{\st'}
-}{
-    \sstep{\stycase{c}{\st}{\sx}{\st_1}{\st_2}}{\argEnv}{\stycase{c}{\st'}{x}{\st_1}{\st_2}}
-}
-
-\inferrule[s-tycase-err]{
-    \serr{\st}{\argEnv}
-}{
-    \serr{\stycase{c}{\st}{\sx}{\st_1}{\st_2}}{\argEnv}
-}
-\end{mathpar}
-\begin{mathpar}\small
-\inferrule[s-tycase-match]{
-    \sval{\sty{c}{\st}}{\argEnv}
-}{
-    \sstep{\stycase{c}{\sty{c}{\st}}{\sx}{\st_1}{\st_2}}{\argEnv}{[\st/x]\st_1}
-}
-
-\inferrule[s-tycase-fail]{
-    \st \neq \sty{c}{\st'}
-}{
-    \sstep{\stycase{c}{\st}{\sx}{\st_1}{\st_2}}{\argEnv}{\st_2}
-}
-\end{mathpar}
-\begin{mathpar}
-\small
-\inferrule[keq-k]{\kvar \in \kDelta}{\keq{\kDelta}{\kvar}}
-
-\inferrule[keq-ind]{\keq{\kDelta, \kvar}{\kappa}}{
-  \keq{\kDelta}{\kmu{\kvar}{\kappa}}}
-
-\inferrule[keq-unit]{ }{\keq{\kDelta}{\kunit}}
-\end{mathpar}\begin{mathpar}
-\small
-\inferrule[keq-prod]{
-  \keq{\kDelta}{\kappa_1}~~~~
-  \keq{\kDelta}{\kappa_2}
-}{
-  \keq{\kDelta}{\kprod{\kappa_1}{\kappa_2}}
-}
-
-\inferrule[keq-sum]{
-  \keq{\kDelta}{\kappa_1}~~~~
-  \keq{\kDelta}{\kappa_2}
-}{
-  \keq{\kDelta}{\ksum{\kappa_1}{\kappa_2}}
-}
-
-\inferrule[keq-ty]{ }{\keq{\kDelta}{\kty}}
-\end{mathpar}
- %We will discuss strategies for fragments that benefit from more sophisticated equivalences in Section \ref{examples}.\todo{do this}
-
-\begin{mathpar}
-
-\inferrule[k-ity-alpha]{ }{\sofkX{\sqity{\alpha}}{\kity}}
-\end{mathpar}
-
-\begin{mathpar}
-\small
-\inferrule[s-ity-lam-step-1]{
-    \sstep{\sqity{\qity_1}}{\argEnv}{\sqity{\qity_1'}}%\sstep{\sqity{\qity_1}}{\argEnv}{\sstep{\sqity{\qity_1'}}}
-}{
-    \sstep{\sqity{\qity_1\times\qity_2}}{\argEnv}{\sqity{\qity_1' \times \qity_2}}
-}
-
-\inferrule[s-ity-lam-step-2]{
-    \sval{\sqity{\qity_1}}{\argEnv}\\
-    \sstep{\sqity{\qity_2}}{\argEnv}{\sqity{\qity_2'}}%\sstep{\sqity{\qity_1}}{\argEnv}{\sstep{\sqity{\qity_1'}}}
-}{
-    \sstep{\sqity{\qity_1\times\qity_2}}{\argEnv}{\sqity{\qity_1 \times \qity_2'}}
-}
-
-\inferrule[s-ity-lam-err-1]{
-    \serr{\sqity{\qity_1}}{\argEnv}
-}{
-    \serr{\sqity{\qity_1\times\qity_2}}{\argEnv}
-}
-
-\inferrule[s-ity-lam-err-2]{
-    \serr{\sqity{\qity_2}}{\argEnv}
-}{
-    \serr{\sqity{\qity_1\times\qity_2}}{\argEnv}
-}
-
-\inferrule[s-ity-lam-v]{
-    \sval{\sqity{\qity_1}}{\argEnv}\\
-    \sval{\sqity{\qity_2}}{\argEnv}
-}{
-    \sval{\sqity{\qity_1\times\qity_2}}{\argEnv}
-}
-
-\inferrule[s-ity-alpha-v]{ }{
-    \sval{\sqity{\alpha}}{\argEnv}
-}
-\end{mathpar}
-\begin{mathpar}
-\small
-\inferrule[k-tycase-parr]{
-    \sofkX{\st}{\kty}\\
-    \sofk{\kDelta}{\kGamma, \sx :: \kty \times \kty}{\Phi}{\st_1}{\kappa}\\
-    \sofkX{\st_2}{\kappa}
-}{
-    \sofkX{\stycase{\rightharpoonup}{\st}{\sx}{\st_1}{\st_2}}{\kappa}
-}
-\end{mathpar}
-\begin{mathpar}
-\small
-\inferrule[s-ity-unquote-step]{
-    \sstep{\st}{\argEnv}{\st'}
-}{
-    \sstep{\sqity{\qtuq{\st}}}{\argEnv}{\sqity{\qtuq{\st'}}}
-}
-
-\inferrule[s-ity-unquote-err]{
-    \serr{\st}{\argEnv}
-}{
-    \serr{\sqity{\qtuq{\st}}}{\argEnv}
-}
-\end{mathpar}
-\begin{mathpar}
-\inferrule[s-ity-trans-step]{
-    \sstep{\st}{\argEnv}{\st'}
-}{
-    \sstep{\sqity{\srep{\st}}}{\argEnv}{\sqity{\srep{\st'}}}
-}
-
-\inferrule[s-ity-trans-err]{
-    \serr{\st}{\argEnv}
-}{
-    \serr{\sqity{\srep{\st}}}{\argEnv}
-}
-\end{mathpar}
-This judgement is defined by the following straightforward rules:
-\begin{mathpar}\small
-\inferrule[tstore-emp]{ }{\emptyset \leadsto \emptyset : \emptyset}
-
-\inferrule[tstore-ext]{\memD \leadsto \delta : \Delta}{(\memD, \st \leftrightarrow \tau/\alpha) \leadsto (\delta, \tau/\alpha) : (\Delta, \alpha)}
-\end{mathpar}
-
-\[\small
-\begin{array}{lll}
-\textbf{Description} ~~~~& \textbf{Concrete Form}~~~~ & \textbf{Desugared Form}\\
-\text{sequences} & (e_1, \ldots, e_n) ~\text{or}~ \texttt{[}e_1, \ldots, e_n\texttt{]} & \eintro{()}{e_1; \ldots; e_n}\\
-\text{labeled sequences} & \{\mathtt{lbl}_1=e_1, \ldots, \mathtt{lbl}_n=e_n\} & \eintro{\texttt{[}\mathtt{lbl}_1, \ldots, \mathtt{lbl}_n\texttt{]}}{e_1; \ldots; e_n}\\
-\text{label application} & \mathtt{lbl}\langle e_1, \ldots, e_n \rangle & \eintro{\mathtt{lbl}}{e_1, \ldots, e_n}\\
-\text{numerals} & n & \eintro{n}{\cdot}\\
-\text{labeled numerals} & n\texttt{lbl} & \eintro{(n, \texttt{lbl})}{\cdot}\\
-\text{strings} & \texttt{"s"} & \eintro{\texttt{"s"}}{\cdot}
-\end{array}
-\]
-
-\begin{mathpar}\small
-\inferrule[s-itm-var-v]{ }{
-    \sval{\sqitm{x}}{\argEnv}
-}
-
-\inferrule[s-itm-lam-step-1]{
-    \sstep{\sqity{\qity}}{\argEnv}{\sqity{\qity'}}
-}{
-    \sstep{\sqitm{\ilam{\qity}{x}{\qitm}}}{\argEnv}{\sqitm{\ilam{\qity'}{x}{\qitm}}}
-}
-
-\inferrule[s-itm-lam-step-2]{
-    \sval{\sqity{\qity}}{\argEnv}\\
-    \sstep{\sqitm{\qitm}}{\argEnv}{\sqitm{\qitm'}}
-}{
-    \sstep{\sqitm{\ilam{\qity}{x}{\qitm}}}{\argEnv}{\sqitm{\ilam{\qity}{x}{\qitm'}}}
-}
-\end{mathpar}
-\begin{mathpar}\small
-\inferrule[s-itm-lam-err-1]{
-    \serr{\sqity{\qity}}{\argEnv}
-}{
-    \serr{\sqitm{\ilam{\qity}{x}{\qitm}}}{\argEnv}
-}
-
-\inferrule[s-itm-lam-err-2]{
-    \serr{\sqitm{\qitm}}{\argEnv}
-}{
-    \serr{\sqitm{\ilam{\qity}{x}{\qitm}}}{\argEnv}
-}
-
-\inferrule[s-itm-lam-v]{
-    \sval{\sqity{\qity}}{\argEnv}\\
-    \sval{\sqitm{\qitm}}{\argEnv}
-}{
-    \sval{\sqitm{\ilam{\qity}{x}{\qitm}}}{\argEnv}
-}
-\end{mathpar}
-\begin{mathpar}\small
-\inferrule[k-itm-unquote]{
-    \sofkX{\st}{\kitm}
-}{
-    \sofkX{\sqitm{\quq{\st}}}{\kitm}
-}
-\end{mathpar}
-
-\begin{mathpar}\small
-\inferrule[s-itm-unquote-step]{
-    \sstep{\st}{\argEnv}{\st'}
-}{
-    \sstep{\sqitm{\quq{\st}}}{\argEnv}{\sqitm{\quq{\st'}}}
-}
-
-\inferrule[s-itm-unquote-err]{
-    \serr{\st}{\argEnv}
-}{
-    \serr{\sqitm{\quq{\st}}}{\argEnv}
-}
-
-\inferrule[s-itm-unquote-elim]{
-    \sval{\sqitm{\qitm}}{\argEnv}
-}{
-    \sstep{\sqitm{\quq{\sqitm{\qitm}}}}{\argEnv}{\sqitm{\qitm}}
-}
-\end{mathpar}
-\begin{mathpar}
-\inferrule[s-ana-step]{
-    \sstep{\st}{\argEnv}{\st'}
-}{
-    \sstep{\sana{n}{\st}}{\argEnv}{\sana{n}{\st'}}
-}
-
-\inferrule[s-ana-err]{
-    \serr{\st}{\argEnv}
-}{
-    \serr{\sana{n}{\st}}{\argEnv}
-}
-
-\inferrule[s-itm-anatrans-step]{
-    \sstep{\st}{\argEnv}{\st'}
-}{
-    \sstep{\sqitm{\anatrans{n}{\st}}}{\argEnv}{\sqitm{\anatrans{n}{\st'}}}
-}
-
-\inferrule[s-itm-anatrans-err]{
-    \serr{\st}{\argEnv}
-}{
-    \serr{\sqitm{\anatrans{n}{\st}}}{\argEnv}
-}
-\end{mathpar}
-, as specified by the judgement $\memG \leadsto \gamma : \Gamma$ defined by the following rules:
-\begin{mathpar}
-\inferrule[ttrs-emp]{ }{\emptyset \leadsto \emptyset : \emptyset}
-
-\inferrule[ttrs-ext]{
-    \memG \leadsto \gamma : \Gamma
-}{
-    (\memG, n : \st \leadsto \iota/x : \tau) \leadsto (\gamma, \iota/x) : (\Gamma, x : \tau)
-}
-\end{mathpar}
-
-\begin{mathpar}\small
-\inferrule[k-itm-lam]{
-    \sofkX{\sqity{\qity}}{\kity}\\
-    \sofkX{\sqitm{\qitm}}{\kitm}
-}{
-    \sofkX{\sqitm{\ilam{\qity}{x}{\qitm}}}{\kitm}
-}
-\end{mathpar} 
-\begin{mathpar}\small
-\inferrule[k-raise]{\kDelta \vdash \kappa}{\sofkX{\sraise{\kappa}}{\kappa}}
-
-\inferrule[s-raise]{ }{\serr{\sraise{\kappa}}{\argEnv}}
-\end{mathpar}
-
-\begin{mathpar}\small
-\inferrule[s-syn-success]{
-    \keyw{nth}[n](\es) = e\\
-    \esynX{e}{\st}{\iota}
-}{
-    \sstep{\ssyn{n}}{\es;\Upsilon;\Phi}{(\st, \sqitm{\syntrans{n}})}
-}
-
-\inferrule[s-syn-fail]{
-    \keyw{nth}[n](\es) = e\\
-    [\Upsilon \vdash_\Phi e \nRightarrow]
-}{
-    \serr{\ssyn{n}}{\es;\Upsilon;\Phi}
-}
-\end{mathpar}
-
-\begin{mathpar}\small
-\inferrule[k-itm-syntrans]{
-    n' < n
-}{
-    \sofkX{\sqitm{\syntrans{n'}}}{\kitm}
-}
-\end{mathpar}
-, e.g. for lambdas:
-\begin{mathpar}\small
-%\inferrule[abs-var]{ }{\edeabs{\tc}{\argEnv}{x}{\memD}{\memG}{x}{\memD}{\memG}}
-%
-\inferrule[abs-lam]{
-    \tdeabs{\Phi}{\tc}{\qity}{\memD}{\ity}{\memD'}\\
-    \edeabs{\tc}{\es;\Upsilon;\Phi}{\qitm}{\memD'}{\memG}{\itm}{\memD''}{\memG'}
-}{
-    \edeabs{\tc}{\es;\Upsilon;\Phi}{\ilam{\qity}{x}{\qitm}}{\memD}{\memG}{\ilam{\tau}{x}{\itm}}{\memG'}{\memD''}
-}
-\end{mathpar}
-\begin{mathpar}\small
-\inferrule[abs-anatrans-stored]{
-    n : \st \leadsto \iota/x : \tau \in \memG
-}{
-    \edeabs{\tc}{\argEnv}{\anatrans{n}{\st}}{\memG}{\memD}{x}{\memG}{\memD}
-}
-
-\inferrule[abs-syntrans-stored]{
-    n : \st \leadsto \iota/x : \tau \in \memG
-}{
-    \edeabs{\tc}{\argEnv}{\syntrans{n}}{\memG}{\memD}{x}{\memG}{\memD}
-}\end{mathpar}
-\begin{mathpar}\small
-\inferrule[k-itm-anatrans]{
-    n' < n\\
-    \sofkX{\st}{\kty}
-}{
-    \sofkX{\sqitm{\anatrans{n'}{\st}}}{\kitm}
-}
-\end{mathpar}
-\begin{mathpar}\small
-\inferrule[abs-syntrans-new]{
-    n \notin \text{dom}(\memG)\\
-    \keyw{nth}[n](\es) = e\\
-    \esynX{e}{\st}{\iota}\\
-    \tdeabs{\Phi}{\tc}{\srep{\st}}{\memD}{\ity}{\memD'}\\
-    (x~\text{fresh})
-}{
-    \edeabs{\tc}{\es;\Upsilon;\Phi}{\syntrans{n}}{\memG}{\memD}{x}{\memG, n : \st \leadsto \iota/x : \tau}{\memD'}
-}
-\end{mathpar}
-
-\begin{mathpar}\small
-\inferrule[etctx-emp]{ }{\vdash_\Phi \emptyset \leadsto \emptyset}
-
-\inferrule[etctx-ext]{\vdash_\Phi \Upsilon \leadsto \Gamma\\\istype{\st}{\Phi}\\\vdash_\Phi \st \leadsto \tau}{\vdash_\Phi \Upsilon, x \Rightarrow \st \leadsto \Gamma, x : \tau}
-\end{mathpar}
-
-\[\small
-\begin{array}{lll}
-\textbf{Description}& \textbf{Concrete Form}~~~~ & \textbf{Desugared Form}\\
-\text{index projection} & e_\text{targ}\texttt{\#}n & \etarg{\opname{idx}}{n}{e_\text{targ}}{\cdot}\\
-\text{label projection} & e_\text{targ}\texttt{\#}\conclbl{lbl} & \etarg{\opname{prj}}{\conclbl{lbl}}{e_\text{targ}}{\cdot}\\
-\text{explicit invocation}~~~~ & e_\text{targ}{\cdot}\opname{op}[\sttmidx](\es) & \etarg{\opname{op}}{\sttmidx}{e_\text{targ}}{\es}\\
-& e_\text{targ}{\cdot}\opname{op}(\es) & \etarg{\opname{op}}{()}{e_\text{targ}}{\es}\\
-& e_\text{targ}{\cdot}\opname{op}(\conclbl{lbl}_1=e_1, \ldots, \conclbl{lbl}_n=e_n) & \keyw{targ}[\opname{op}; \texttt{[}\conclbl{lbl}_1, \ldots, \conclbl{lbl}_n\texttt{]}](e_\text{targ}; \\
-& & \quad e_1; \ldots; e_n)\\
-\text{labeled case analysis} & e_\text{targ}{\cdot}\opname{case}\,\{ & \keyw{targ}[{\opname{case}}; {\texttt{[}\st_1, \ldots, \st_n\texttt{]}}]({e_\text{targ}}; \\
- & ~|~~\st_1\langle x_1, \ldots, x_k \rangle \Rightarrow e_1 & \quad \eanalam{x_1}{\ldots \eanalam{x_k}{e_1}};\\
- & ~|~~\ldots & \quad \ldots;\\ 
- & ~|~~\st_n\langle x_1, \ldots, x_k \rangle \Rightarrow e_n\} & \quad \eanalam{x_1}{\ldots \eanalam{x_k}{e_n}})
-\end{array}
-\]
-
- For example, 
-\begin{mathpar}
-\small
-\inferrule[abs-prod]{
-    \tdeabs{\Phi}{\tc}{\qity_1}{\memD}{\ity_1}{\memD'}\\
-    \tdeabs{\Phi}{\tc}{\qity_2}{\memD'}{\ity_2}{\memD''}
-}{
-    \tdeabs{\Phi}{\tc}{\qity_1 \times \qity_2}{\memD}{\ity_1 \times \ity_2}{\memD''}
-}
-\end{mathpar}
-
-The argument interfaces that populate the list provided to opcon definitions is derived from the argument list by the judgement $\keyw{args}(\es)=_n \stx{args}$, defined as follows:
-\begin{mathpar}\small
-\inferrule[args-z]{ }{
-    \keyw{args}(\cdot)=_0 \skap{\karg}{\svar{nil}}
-}
-
-\inferrule[args-s]{
-    \keyw{args}(\es)=_{n} \st
-}{
-    \keyw{args}(\es;e)=_{n+1} \skap{\karg}{\svar{rcons}}~\st~(\slam{\kty}{\svar{ty}}{\sana{n}{\svar{ty}}},\slam{\kunit}{\_}{\ssyn{n}})
-}
-\end{mathpar}
-We assume that the definitions of the standard helper functions $\small\svar{nil} :: \kforall{\kalpha}{\klist{\kalpha}}$ and $\small\svar{rcons} :: \kforall{\kalpha}{\klist{\kalpha} \rightarrow \kalpha \rightarrow \klist{\kalpha}}$, which adds an item to the end of a list, have been substituted into these rules.
-The result is that the $n$th element of the argument interface list simply wraps the static terms $\sana{n}{\st}$ and $\ssyn{n}$.\end{document}
+\end{document}

### 2014-11-02 att-pldi15/att-pldi15.tex
compressing space a bit

@@
 Modular reasoning will be based on holding the type translation of $\st$ abstract ``outside'' the tycon.
+
+ External typing contexts $\Upsilon$ map variables to types, so the type translation  judgement induces judgement  $\vdash_\Phi \Upsilon \leadsto \Gamma$.
 
@@
 The semantics also directly follow \cite{pfpl}, so we  omit the details of this core here.
-Only three new  kinds will be needed: $\kty$ (Sec.
+Only three new  kinds are needed for the SL to serve its role as the language used to  statically compute over types and type and term translations: $\kty$ (Sec.
 \ref{types}), $\kity$ (Sec.
@@
 \caption{Syntax of tycons. Metavariables $\tc$ and $\opname{op}$ range over user-defined tycon and opcon names, respectively, and $m$ ranges over natural numbers.}
-\label{syntax-TC}
+\label{syntax-TC}\vspace{-5px}
 \end{figure}
@@
 \caption{Kinding rules for types, which take the form $\sty{c}{\sttyidx}$ where $c$ is a tycon and $\sttyidx$ is the type index.}
-\label{fig:types}
+\label{fig:types}\vspace{-5px}
 \end{figure*}
@@
 For example, the kinding rule for when $c$ is user-defined is below.
+\vspace{-5px}
 \begin{mathpar}
@@
 
- %The dynamics (see supplement) are again straightforward, consistent with this intuition.% We will see an example of its use in Sec. \ref{sec:targops}.
+The dynamics (see supplement) are straightforwardly consistent with the intuition above.% We will see an example of its use in Sec. \ref{sec:targops}.
 \begin{figure}\hfill \fbox{$\vdash \Phi$}\vspace{-25px}\begin{mathpar}
@@
 
-\subsection{Typing Contexts}
-External typing contexts contain types, so we can now also define a translation judgement  $\vdash_\Phi \Upsilon \leadsto \Gamma$ (see supplement).
 %\vspace{-5px}
@@
 We chose not to formalize these for simplicity, and to demonstrate this general technique.
-An analagous technique could be used to implement record types by requiring that the index be sorted.%, leaving indices where the labels did not appear sorted uninhabited.
+%An analagous technique could be used to implement record types by requiring that the index be sorted.%, leaving indices where the labels did not appear sorted uninhabited.
 In both cases, we could provide a static helper function of kind $\klist{\klbl\times\kty}\rightarrow \kty$ that checked well-formedness immediately before constructing the requested type.
@@
 
-\paragraph{Other Intro} The rule (ana-intro-other) is used to introduce terms of a type constructed by an ``other'' tycon.
+The rule (ana-intro-other) is used to introduce terms of a type constructed by an ``other'' tycon.
 The term index, rather than the tycon context, directly specifies the static function that maps the arguments to a translation.
@@
 \]
-The first line checks that exactly two arguments, including the target argument, were passed in via helper function $\svar{arity2}$.
+The helper function $\svar{arity2}$ checks that exactly two arguments, including the target argument.
 We then request synthesis of both arguments.
 We can ignore the type synthesized by the first because by definition it is a regular string type with type index $\svar{tyidx}$.
-We case analyze the second against $\tcvar{rstr}$, extracting its index regular expression if so and failing if not.
-We finally synthesize the resulting regular string type, using the helper function $\svar{rxconcat} :: \krx \rightarrow \krx \rightarrow \krx$ which generates the concatenated regular expression, and the translation, using an internal helper function $sconcat : \keyw{str} \rightarrow \keyw{str} \rightarrow \keyw{str}$, the translational internal term for which we assume has been substituted in directly above.
+We case analyze the second against $\tcvar{rstr}$, extracting its index regular expression if so and raising an error if not.
+We then synthesize the resulting regular string type, using the helper function $\svar{rxconcat} :: \krx \rightarrow \krx \rightarrow \krx$ which generates the synthesized type index, and finally the translation, using an internal helper function $sconcat : \keyw{str} \rightarrow \keyw{str} \rightarrow \keyw{str}$, the translational term for which we assume has been substituted in directly above.
 
@@
 
-\paragraph{Other Targeted Ops} Like (ana-intro-other), rule (syn-targ-other) is used when the target synthesizes an ``other'' type.
+Like (ana-intro-other), rule (syn-targ-other) is used when the target synthesizes an ``other'' type.
 The mapping from the arguments to a type and translation is given directly in the term index (the op name is ignored).
@@
 Such a calculus and compiler specification was given by Fulton et al.
-\cite{sanitation-psp14} and extending this to our  setting requires showing only that the opcon definitions in $\tcvar{rstr}$ adequately capture these specifications using standard program correctness techniques for the SL, which is a simply-typed functional language.
+\cite{sanitation-psp14} and extending this to our  setting requires showing only that the opcon definitions in $\tcvar{rstr}$ adequately capture these specifications using standard program correctness techniques for the SL, a simply-typed functional language.
 The only ``wrinkle'' is that the rule (syn-targ-other) can synthesize a regular string type.
-But if so, the abstracted translation will be checked against $\tau_\text{abs}=\alpha$, so only strings that could otherwise arise can result, a  consequence of the standard abstraction theorem for the IL.\todo{citation} \end{proof-sketch}
+But if so, the abstracted translation will be checked against $\tau_\text{abs}=\alpha$, so only strings that could otherwise arise can result, by the abstraction theorem for the IL.\todo{citation} \end{proof-sketch}
 

### 2014-11-02 att-pldi15/att-pldi15.tex
More edits

@@
 
-In this paper, we begin from first principles with a core calculus, @$\lambda$, specified like many full-scale languages: as a bidirectionally typed translation semantics.
+In this paper, we begin from first principles by specifying a language, @$\lambda$, in the style of many full-scale languages: as a bidirectionally typed translation semantics.
 Only the $\rightarrow$ type constructor (tycon) is built in; all other external tycons (we show a variant on record types and constrained string types) are defined by extending a \emph{tycon context}.
-Each tycon defines the semantics of its associated term-level operators (e.g.
-record projection) using functions written in a static language where types and translations are values.
-The semantics provide  strong metatheoretic guarantees, notably \emph{type safety} and \emph{conservativity}: that all \emph{tycon  invariants} will be conserved under extension.
-Mechanized proofs are not needed: problems are caught during typechecking by lifting typed compilation techniques into the semantics and enforcing barriers around tycons using type abstraction, the same principle that underlies reasoning with ML-style modules.
+Each tycon defines the semantics of its  term-level operators (e.g.
+record projection) using a static language where types and translations are values.
+The semantics includes a simple check that leads to powerful metatheoretic guarantees, notably \emph{type safety} and \emph{conservativity}: that  \emph{tycon invariants} will always be conserved under extension.
+Type system providers need not provide mechanized proofs.
+Instead, the check enforces a \emph{translation independence} property via type abstraction, the same principle underlying ML-style module systems.
 \end{abstract}
@@
 \section{Introduction}\label{intro}
-Typed programming languages are most often described as being composed of \emph{fragments}, each contributing to the language's concrete syntax, abstract syntax, static semantics and dynamic semantics.
-In his textbook, Harper organizes fragments around type constructors, describing each in a different chapter \cite{pfpl}.
-Languages are then identified by a set of type constructors, e.g.
-$\mathcal{L}\{{\rightharpoonup}\,{\forall}\,{\mu}\,{\keyw{1}}\,{\times}\,{+}\}$ is the language that includes partial function types, polymorphic types, recursive types, nullary and binary product types and binary sum types (its syntax is shown in Figure \ref{syntax-IL}, discussed below).
-Another related approach is to describe fragments with a simple calculus having a ``catch-all'' constant and base type to stand notionally for all other terms and types that may also be included in some future complete language.
+Typed programming languages are most often described as being composed from \emph{fragments}, each contributing to the language's concrete syntax, abstract syntax, static semantics and dynamic semantics.
+In his textbook, Harper organizes such fragments around type constructors, describing each in a different chapter \cite{pfpl}.
+Languages are then identified by a set of these type constructors, e.g.
+$\mathcal{L}\{{\rightharpoonup}\,{\forall}\,{\mu}\,{\keyw{1}}\,{\times}\,{+}\}$ is the language of partial function types, polymorphic types, recursive types, nullary and binary product types and binary sum types (its syntax is shown in Figure \ref{syntax-IL}, discussed below).
+In the literature, researchers often describe fragments using a simple calculus having a ``catch-all'' constant and base type to stand notionally for all other terms and types that may also be included in some future complete language.
 
@@
 Universality properties (e.g.
-``Turing-completeness'') can usually be invoked to guarantee that an embedding that preserves a desirable fragment's dynamics is possible, but an isomorphic embedding must also preserve the static semantics and, if defined, performance bounds specified using a cost semantics.
+``Turing-completeness'') are often invoked to guarantee that an embedding that preserves a desirable fragment's dynamics can be constructed, but an isomorphic embedding must also preserve the static semantics and, if defined, performance bounds specified using a cost semantics.
 This is not always possible.
@@
 %Indeed, the proliferation of language dialects constructed for this reason might be taken as an evidence that the core language is not as ``general'', when considered comprehensively, as might be hoped.
-Each time a fragment like this is needed, a new \emph{dialect} of a language must be constructed. Within the ML lineage, for example, dialects abound:
+Each time a fragment must be exposed directly, rather than as a library, a new \emph{dialect} of a language must be constructed.
+Within the ML lineage, for example, dialects that go beyond ``core ML'' abound:
 %Reynolds, in a remark that recalls the ``Turing tarpit'' of Perlis \cite{Perl82a}, summarizes the issue \cite{Reynolds94anintroduction}: 
@@
 These problems do not arise for a fragment that can be expressed as an isomorphic embedding (i.e.
-as a library) because modern \emph{module systems} can enforce abstraction barriers that ensure that the isomorphism need only be established in the ``closed world'' of the module.
+as a library) because modern \emph{module systems} can enforce abstraction barriers that ensure that the isomorphism needs only to be established in the ``closed world'' of the module.
 For example, a module defining sets in ML can hold the representation of sets abstract, ensuring that any invariants maintained by the functions in the module (e.g.
-uniqueness, if using a list representation) will hold no matter which other modules are separately in use by a client.\todo{citation?} %Mechanisms that can help decrease the complexity of an embedding without violating abstraction barriers are thus valuable, and we will lead into our work in Sec.
+uniqueness, if using a list representation) will hold no matter which other modules are separately in use by a client.
+Other langauges offer similar facilities, sometimes called \emph{abstract data types}.\todo{citation?} %Mechanisms that can help decrease the complexity of an embedding without violating abstraction barriers are thus valuable, and we will lead into our work in Sec.
 \ref{prior-work} by summarizing them.
@@
 \ref{types}.
-We describe how tycons control the semantics of their associated term-level operator constructors (opcons) in Sec.
+We describe how tycons control the semantics of their term-level operator constructors (opcons) in Sec.
 \ref{external-terms}.
@@
 in the ``open world'').
-Interestingly, type system providers need not necessarily provide mechanized proofs to maintain these guarantees.
+Interestingly, type system providers need not provide mechanized proofs to maintain these guarantees.
 Instead, the approach we take relies on type abstraction in the internal language.
-As a result, we are able to borrow the same results that underly modular reasoning in simply-typed languages like ML to reason modularly about language fragments.
+As a result, we are able to borrow the same results that underly modular reasoning in simply-typed languages like ML to reason modularly about typed language fragments.
 We conclude by discussing  related and future work in Sec.
@@
 The internal language is purposely kept small, e.g.
-defining only simple products, to simplify metatheoretic reasoning and compilation.
+defining only binary products, to simplify metatheoretic reasoning and compilation.
 The EL then specifies various useful higher-level constructs, e.g.
@@
 In @$\lambda$, the EL builds in only function types.
-All other external constructs are defined in the \emph{tycon context}, described in the Sec.
+All other external constructs are defined in the \emph{tycon context}, described starting in the Sec.
 \ref{types}.
@@
 
- External typing contexts $\Upsilon$ map variables to types, so the type translation  judgement induces judgement  $\vdash_\Phi \Upsilon \leadsto \Gamma$.
+External typing contexts $\Upsilon$ map variables to types, so we also need the judgement $\vdash_\Phi \Upsilon \leadsto \Gamma$, where $\Gamma$ maps variables to the corresponding type translation.
+In examples, we omit leading $\emptyset$, used as the base case for finite mappings, and $\cdot$, used as the base case for finite sequences, for concision.
 
@@
 %The typing and type formation contexts obey standard structural properties (i.e. weakening, exchange and contraction). 
-In examples, we will omit leading $\emptyset$, used as the base case for finite mappings, and $\cdot$, used as the base case for finite sequences (e.g. writing $\Gamma_\text{test} := x : \tau$).
-
 The internal dynamics are specified as a structural operational semantics with a stepping judgement {\small $\iota \mapsto \iota\moutput$} and a value judgement {$\iota~\mathtt{val}$}.
 The multi-step judgement $\iota \mapsto^{*} \iota\moutput$ is the reflexive, transitive closure of the stepping judgement and the evaluation judgement $\iota \Downarrow \iota'$ is defined iff $\iota \mapsto^{*} \iota'$ and $\iota'~\mathtt{val}$.
-Both the static and dynamic semantics of the IL can be found in any standard textbook covering typed lambda calculi (we directly follow \cite{pfpl}), so we assume familiarity and key metatheoretic properties.
+Both the static and dynamic semantics of the IL can be found in any standard textbook covering typed lambda calculi (we directly follow \cite{pfpl}), so we assume familiarity.
 
@@
 The semantics also directly follow \cite{pfpl}, so we  omit the details of this core here.
-Only three new  kinds are needed for the SL to serve its role as the language used to  statically compute over types and type and term translations: $\kty$ (Sec.
+Only three new  kinds are needed for the SL to serve its role as the language used to   control typechecking and translation of the EL: $\kty$ (Sec.
 \ref{types}), $\kity$ (Sec.
@@
 
-All user-defined tycons must be defined in the \emph{tycon context}, $\Phi$, which is simply a list of tycon definitions.
-Each tycon defines a name, $\tc$, a \emph{tycon structure}, $\theta$, and a \emph{tycon signature}, $\psi$.
-We will return to the tycon structure below.
+All user-defined tycons $\tc$ must be defined in the \emph{tycon context}, $\Phi$, which is simply a list of tycon definitions, $\tcdef{\tc}{\psi}{\theta}$, where $\theta$ is the \emph{tycon structure} and $\psi$ is the \emph{tycon signature}.
+We will return to tycon structures below.
 Tycon signatures have the form $\tcsig{\ktyidx}{\chi}$, where $\ktyidx$ is the tycon's index kind and $\chi$ is the \emph{opcon signature}, which we   discuss in Sec.
@@
 It has index kind $\krx$, which classifies static regular expression patterns (defined as an inductive sum kind in the usual way).
-Types constructed by $\tcvar{rstr}$ will classify \emph{regular strings}, which are statically known to be in the regular language specified by the type index \cite{sanitation-psp14}.
+Types constructed by $\tcvar{rstr}$ will classify \emph{regular strings}, which statically obey the invariant that they are in the regular language specified by the type index \cite{sanitation-psp14}.
 For example, $\stx{title} := \sty{\tcvar{rstr}}{\concrx{.+}}$ will classify non-empty strings and $\stx{conf} := \sty{\tcvar{rstr}}{\concrx{[A-Z]+ \digit\digit\digit\digit}}$ will classify conference names.
 The type indices are  here written using standard concrete syntax for concision; recent work has specified how to define type-specific (or here, kind-specific) syntax like this composably in libraries \cite{TSLs}.
-We define a tycon context containing only the definition of $\tcvar{rstr}$, $\Phi_\text{rstr} := \tcdef{\tcvar{rstr}}{\tcsig{\krx}{\chi_\text{rstr}}}{\theta_\text{rstr}}$.
+We define the tycon context containing only the definition of $\tcvar{rstr}$, $\Phi_\text{rstr} := \tcdef{\tcvar{rstr}}{\tcsig{\krx}{\chi_\text{rstr}}}{\theta_\text{rstr}}$.
 
@@
 Each tycon in the tycon context {computes} translations for the types it constructs as a function of each type's index by specifying a \emph{translation schema} in the {tycon structure}, $\theta$.
-A tycon with index kind $\ktyidx$ defines a translation schema of kind $\ktyidx \rightarrow \kity$, checked by (tcc-ext).
+A tycon with index kind $\ktyidx$ must define a translation schema of kind $\ktyidx \rightarrow \kity$, checked by (tcc-ext).
 
@@
 Each form in the syntax for internal types, $\tau$,  corresponds to a form in the syntax of translational internal types, $\qity$.
-For example, our translation schema for $\tcvar{rstr}$  simply chooses to ignore the type index and represent all regular strings internally as strings, of internal type abbreviated $\keyw{str}$.
-We abbreviate the corresponding translational internal type $\hat{\keyw{str}}$ and define the translation schema as $\small\theta_\text{rstr} := \tcstruct{\slam{\krx}{\svar{tyidx}}{\sqity{\hat{\keyw{str}}}}}{\omega_\text{rstr}}$.
+For example, our schema for $\tcvar{rstr}$ will simply choose to ignore the type index and translate regular strings to strings, of internal type abbreviated $\keyw{str}$.
+We abbreviate the corresponding translational internal type $\hat{\keyw{str}}$ and define $\small\theta_\text{rstr} := \tcstruct{\slam{\krx}{\svar{tyidx}}{\sqity{\hat{\keyw{str}}}}}{\omega_\text{rstr}}$.
 
@@
 
-We choose a translation schema for $\tcvar{lprod}$ that generates nested binary product types by recursing over the type index and referring to the translations of the types therein (though this is not the only workable choice, e.g.
+These are used in the schema for $\tcvar{lprod}$, which generates nested binary product types by recursing over the type index and referring to the translations of the types therein (this is not the only viable choice, e.g.
 we could also have used a list).
@@
 \subsubsection{Selective Type Translation Abstraction}\label{sec:selective-type-translation-abstraction}
-References to translations of other types are maintained in values of kind $\kity$ like this to  allow us to selectively hold them abstract, which will be the key to our main results.
+References to type translations are maintained in values of kind $\kity$ like this to  allow us to selectively hold them abstract, which will be the key to our main results.
 This can be thought of as analagous to the process in ML by which the true identity of an abstract type in a module is held abstract outside the module until after typechecking.
-The judgement $\small\tdeabs{\Phi}{c}{\qity}{\memD}{\ity\moutput}{\memD\moutput}$ relates a normalized translational internal type $\qity$ to an internal type $\tau$, called a \emph{selectively abstracted type translation} because references to translations of types constructed by a tycon other than the ``delegated tycon'', $c$, are replaced by a corresponding type variables, $\alpha$.
-The \emph{type translation store} $\memD ::= \emptyset ~|~ \memD, \st \leftrightarrow \ity/\alpha$ maintains the corresponding between types, their actual translations and the unique type variables which appear in their place.
-For example, $\small\tdeabs{\Phi_\text{rstr}\Phi_\text{lprod}}{\tcvar{lprod}}{\qity_\text{paper/trans}}{\emptyset}{\tau_\text{paper/abs}}{\memD_\text{paper/abs}}$
+The judgement $\small\tdeabs{\Phi}{c}{\qity}{\memD}{\ity\moutput}{\memD\moutput}$ relates a normalized translational internal type $\qity$ to an internal type $\tau$, called a \emph{selectively abstracted type translation} because references to translations of types constructed by a tycon other than the ``delegated tycon'', $c$, are replaced by a corresponding type variable, $\alpha$.
+The \emph{type translation store} $\memD ::= \emptyset ~|~ \memD, \st \leftrightarrow \ity/\alpha$ maintains the correspondence between types, their actual translations and the unique type variables which appear in their place, e.g.
+$\small\tdeabs{\Phi_\text{rstr}\Phi_\text{lprod}}{\tcvar{lprod}}{\qity_\text{paper/trans}}{\emptyset}{\tau_\text{paper/abs}}{\memD_\text{paper/abs}}$
 where $\small\tau_\text{paper/abs} := \alpha_1 \times (\alpha_2 \times \iunit)$ and 
@@
 \begin{mathpar}\small
-\inferrule[conc-ty-trans]{
+\inferrule[ty-trans]{
     \istype{\st}{\Phi}\\
@@
 
-The rules for the selective type abstraction judgement are straightforward.
+The rules for the selective type translation abstraction judgement are straightforward.
 We first recurse generically over sub-terms of $\qity$ until sub-terms of form $\srep{\st}$ are encountered (see supplement).
@@
 
-Having established how types are constructed, and how they determine selectively abstract and from there actual type translations, we can finally give the typing and translation rules for external terms, shown in Figure \ref{typing}.
+Having established how types are constructed, and how they determine selectively abstracted and from there actual type translations, we can finally give the typing and translation rules for external terms, shown in Figure \ref{typing}.
 
@@
 This is a static function that is applied, in the seventh premise, to determine whether the term is well-typed, raising an error if not or determining the translation of the term if so.
-The function has access to the type index, the term index and an interface to the list of arguments, and its kind is checked by the  judgement $\vdash_\Phi \omega \sim \psi$, which appeared as the final premise of the rule (tcc-ext) and is defined in Figure \ref{ocstruct}.
+The function has access to the type index, the term index and an interface to the list of arguments, and is kind is checked by the  judgement $\vdash_\Phi \omega \sim \psi$, which appeared as the final premise of the rule (tcc-ext) and is defined in Figure \ref{ocstruct}.
 
@@
 The rest of this opcon definition folds over the three lists provided as input: the list mapping labels to types provided as the type index, the list of labels provided as the term index, and the list of argument interfaces.
-We assume a straightforward helper function, $\svar{listfold3}$, that raises an error if the three lists are not of the same length.
+We assume a straightforward helper function, $\svar{listrec3}$, that raises an error if the three lists are not of the same length.
 The base case is the translational empty product.
@@
 Then, we request type analysis of the corresponding argument, $\svar{rowarg}$,  against the type in the type index, $\svar{rowty}$, by writing $\svar{ana}~\svar{rowarg}~\svar{rowty}$.
-Here, $\svar{ana}$ is a helper function defined below that produces a translational internal term  of the form $\sqitm{\anatrans{n}{\st}}$, where $\st$ is the value of $\svar{rowty}$, if the argument is well-typed, and raises an error if not.
+Here, $\svar{ana}$ is a helper function defined below that produces a translational internal term  of the form $\sqitm{\anatrans{n}{\st}}$, where $n$ is the position of $\svar{rowarg}$ in $\svar{args}$ and $\st$ is the value of $\svar{rowty}$, if the argument is well-typed, and raises an error if not.
 The final line constructs a nested tuple based on this translation and the recursive result.
@@
 
-The argument interface list generated by the judgement $\keyw{args}(n)=\stx{args}$, where $n$ is the length of the argument list, written $|\es|=n$, is then simply a static value of kind $\kargs$ of length $n$ where the $i$th entry is $(\slam{\kty}{\svar{ty}}{\sana{i}{\svar{ty}}}, \slam{\kunit}{\_}{\ssyn{i}})$.
+The argument interface list is is then simply a static value of kind $\kargs$ of length $n$ where the $i$th entry is $(\slam{\kty}{\svar{ty}}{\sana{i}{\svar{ty}}}, \slam{\kunit}{\_}{\ssyn{i}})$.
+It is generated by the judgement $\keyw{args}(n)=\stx{args}$, where $n$ is the length of the argument list, written $|\es|=n$.
 
@@
 The rule (ocstruct-intro) ruled out writing either of these forms explicitly in an opcon definition by checking against the bound $n=0$.
-This is to prevent out-of-bounds errors -- tycon providers do not write these forms directly, but only access them via the argument interface list, guaranteed to have the correct length.
-This is  safe because:
+This is to prevent out-of-bounds errors -- tycon providers do not write these forms directly, only accessing them via the argument interface list, guaranteed to have the correct length.
+This is  safe because we can weaken the upper bound:
 \begin{lemma}
@@
 
-\paragraph{Selectively Abstracted Term Translations} The reason for this  is again because we will hold these abstract by replacing them with variables.
-The judgement $\edeabs{c}{\argEnv}{\qitm}{\memD}{\memG}{\itm\moutput}{\memD\moutput}{\memG\moutput}$, appearing as the eighth premise of (ana-intro), relates a translational internal term $\qitm$ to an internal term $\itm$ called its \emph{selectively abstracted term translation}, where all references to the translation of an argument (having any type) are replaced with a unique variable and all references to the translation of a type constructed by a user-defined tycon other than the ``delegated tycon'' $\tc$ are replaced with an abstract type variable as described in Sec.
+\paragraph{Selectively Abstracted Term Translations} The reason  is again because we will hold these abstract by replacing them with variables.
+The judgement $\edeabs{c}{\argEnv}{\qitm}{\memD}{\memG}{\itm\moutput}{\memD\moutput}{\memG\moutput}$, appearing as the eighth premise of (ana-intro), relates a translational internal term $\qitm$ to an internal term $\itm$ called a \emph{selectively abstracted term translation}, because all references to the translation of an argument (having any type) are replaced with a corresponding variable and all references to the translation of a type constructed by a user-defined tycon other than the ``delegated tycon'' $c$ are replaced with a corresponding abstract type variable as described in Sec.
 \ref{sec:selective-type-translation-abstraction}.
-The type translation store, $\memD$, discussed previously, and term translation store, $\memG$, track these mappings.
-Term translation stores have the following syntax: $\memG ::= \emptyset ~|~ \memG, n : \st \leadsto \iota/x : \tau$.
+The type translation store, $\memD$, discussed previously, and term translation store, $\memG$, track these correspondences.
+Term translation stores have  syntax $\memG ::= \emptyset ~|~ \memG, n : \st \leadsto \iota/x : \tau$.
 Each entry can be read ``argument $n$ having type $\st$ and translation $\iota$ appears as variable $x$ with type $\tau$''.
-For example, \[\small\edeabs{\tcvar{lprod}}{(title, \texttt{"EXMPL 2015"}),\Upsilon,\Phi_\text{rstr}\Phi_\text{lprod}}{\qitm_\text{ex}}{\emptyset}{\emptyset}{\iota_\text{ex/abs}}{\memD_\text{ex/abs}}{\memG_\text{ex/abs}}\] where $\itm_\text{ex/abs} := (x_0, (x_1, ()))$, the term translation store is $\memG_\text{ex/abs} := 0 : \stx{title} \leadsto title/x_0 : \alpha_0, 1 : \stx{conf} \leadsto \texttt{"EXMPL 2015"}_\text{IL}/x_1 : \alpha_1$ and the type translation store $\memD_\text{ex/abs}$ is alpha-equivalent to $\memD_\text{paper/abs}$ from Sec.
+For example, \[\small\edeabs{\tcvar{lprod}}{(title; \texttt{"EXMPL 2015"});\Upsilon;\Phi_\text{rstr}\Phi_\text{lprod}}{\qitm_\text{ex}}{\emptyset}{\emptyset}{\iota_\text{ex/abs}}{\memD_\text{ex/abs}}{\memG_\text{ex/abs}}\] where $\itm_\text{ex/abs} := (x_0, (x_1, ()))$, the term translation store is $\memG_\text{ex/abs} := 0 : \stx{title} \leadsto title/x_0 : \alpha_0, 1 : \stx{conf} \leadsto \texttt{"EXMPL 2015"}_\text{IL}/x_1 : \alpha_1$ and the type translation store $\memD_\text{ex/abs}$ is alpha-equivalent to $\memD_\text{paper/abs}$ from Sec.
 \ref{sec:selective-type-translation-abstraction}.
 
-The rules for the abstracted term translation judgement follow recursively for shared forms just like those for the abstracted type translation judgement (see supplement).
-The rule for references to argument translations derived via type analysis operates as follows ($\syntrans{n}$ is analagous):
+The selectively abstracted term translation judgement proceeds recursively along shared forms, like the selectively abstracted type translation judgement.
+The key rule for references to argument translations derived via analysis is below ($\syntrans{n}$ is analagous; see supplement):
 \begin{mathpar}\small
@@
 
-The tenth premise of (ana-intro) determines an abstract type translation for the type provided for analysis, in this case $\small\tau_\text{ex/abs} := \alpha_0 \times (\alpha_1 \times \kunit)$ (alpha-equivalent to $\tau_\text{paper/abs}$ in Sec.
+The tenth premise of (ana-intro) determines a selectively abstracted type translation for the type provided for analysis, also using the tycon of the type provided for analysis as the delegated tycon and starting with the same store to ensure that equal types have equal type variables.
+In this case $\small\tau_\text{ex/abs} := \alpha_0 \times (\alpha_1 \times \kunit)$ (alpha-equivalent to $\tau_\text{paper/abs}$ in Sec.
 \ref{sec:selective-type-translation-abstraction}), and the eleventh premise extracts a type substitution, $\delta_\text{ex/abs}$, and type formation context, $\Delta_\text{ex/abs}$, from $\memD_\text{ex/abs}$, again equivalent to $\delta_\text{conf/abs}$ and $\Delta_\text{conf/abs}$ from Sec.
@@
 The opcons associated with $\tcvar{rstr}$ are taken directly from Fulton et al.'s specification of regular string types \cite{sanitation-psp14}, with the exception of $\opname{case}$, which generalizes case analysis as defined there to arbitrary string patterns, which we discuss in the supplement.
-The opcons associated with $\tcvar{lprod}$ are also straightforward: $\opname{prj}$ projects out the row with the provided label, $\opname{conc}$ concatenates two labeled products (updating common fields with the value of the right argument), and $\opname{drop}$ creates a new labeled product from the target with some fields dropped.
+The opcons associated with $\tcvar{lprod}$ are also straightforward: $\opname{prj}$ projects out the row with the provided label, $\opname{conc}$ concatenates two labeled products (updating common rows with the value from the right argument), and $\opname{drop}$ creates a new labeled product from the target with some rows dropped.
 Note that both regular strings and labeled products can define concatenation without conflict.
@@
 
-The fifth premise of (syn-targ) extracts the targeted opcon definition of $\opname{op}$ from the opcon structure, $\omega$.
+The fifth premise of (syn-targ) extracts the \emph{targeted opcon definition} of $\opname{op}$ from the opcon structure, $\omega$.
 Like the intro opcon definition, this is a static function that generates a translational internal term on the basis of the target tycon's type index, the term index and an argument interface list.
@@
 \]
-The helper function $\svar{arity2}$ checks that exactly two arguments, including the target argument.
+The helper function $\svar{arity2}$ checks that two arguments, including the target argument, were provided.
 We then request synthesis of both arguments.
@@
 We case analyze the second against $\tcvar{rstr}$, extracting its index regular expression if so and raising an error if not.
-We then synthesize the resulting regular string type, using the helper function $\svar{rxconcat} :: \krx \rightarrow \krx \rightarrow \krx$ which generates the synthesized type index, and finally the translation, using an internal helper function $sconcat : \keyw{str} \rightarrow \keyw{str} \rightarrow \keyw{str}$, the translational term for which we assume has been substituted in directly above.
-
-The remaining premises of (syn-targ) are analagous to the corresponding premises in (ana-intro), with the only difference being that we check the abstract term translation against the abstract type translation of the synthesized type.
+We then synthesize the resulting regular string type, using the helper function $\svar{rxconcat} :: \krx \rightarrow \krx \rightarrow \krx$ which generates the synthesized type index, and finally the translation, using an internal helper function $sconcat : \keyw{str} \rightarrow \keyw{str} \rightarrow \keyw{str}$, the translational term for which we assume has been substituted in directly.
+
+The remaining premises of (syn-targ) are analagous to the corresponding premises in (ana-intro), with the only difference being that we check the selectively abstract term translation against the selectively abstracted type translation of the synthesized type, but the delegated tycon is that of the type synthesized by the target argument.
 
@@
 \end{theorem}
-The case in the proof of Theorem \ref{thm:static-preservation} for $\ssyn{n}$ requires that the following theorem be simultaneously defined.
+The case in the proof of Theorem \ref{thm:static-preservation} for $\ssyn{n}$ requires that the following theorem be mutually defined.
 The mutual induction is well-founded because the total number of argument lists in the terms being considered decreases.
@@
 The interesting cases are (ana-intro), (ana-intro-other), (syn-trans) and (syn-trans-other); the latter two arise via subsumption.
-The result follows directly from the final premise of each rule, combined with lemmas that state that if $\memD \leadsto \delta : \Delta_\text{abs}$ and $\memG \leadsto \gamma : \Gamma_\text{abs}$ then $\vdash \delta : \Delta$ and $\Delta_\text{abs}~\Gamma \vdash \gamma : \Gamma_\text{abs}$, i.e.
+The result follows directly from the final premise of each rule, combined with lemmas that state that if $\memD \leadsto \delta : \Delta_\text{abs}$ and $\memG \leadsto \gamma : \Gamma_\text{abs}$ then $\vdash \delta : \Delta_\text{abs}$ and $\Delta_\text{abs}~\Gamma \vdash \gamma : \Gamma_\text{abs}$, i.e.
 that all the variables in $\Delta_\text{abs}$ and $\Gamma_\text{abs}$ have well-formed/well-typed substitutions in $\delta$ and $\gamma$,  and thus that applying them in the conclusions of the rules gives a well-typed term.
@@
 
-\paragraph{Stability}
+\paragraph{Stability}\todo{if space, bring unicity back}
 Extending the tycon context does not change the meaning of any terms that were previously well-typed.
@@
 Such a calculus and compiler specification was given by Fulton et al.
-\cite{sanitation-psp14} and extending this to our  setting requires showing only that the opcon definitions in $\tcvar{rstr}$ adequately capture these specifications using standard program correctness techniques for the SL, a simply-typed functional language.
-The only ``wrinkle'' is that the rule (syn-targ-other) can synthesize a regular string type.
-But if so, the abstracted translation will be checked against $\tau_\text{abs}=\alpha$, so only strings that could otherwise arise can result, by the abstraction theorem for the IL.\todo{citation} \end{proof-sketch}
+\cite{sanitation-psp14}, so we must simply show that the opcon definitions in $\tcvar{rstr}$ adequately capture these specifications using standard program correctness techniques for the SL, a simply-typed functional language.
+The only ``twist'' is that the rule (syn-targ-other) can synthesize a regular string type.
+But if so, the abstracted term translation will be checked against $\tau_\text{abs}=\alpha$, so only strings that could otherwise arise can result, by the standard abstraction theorem for the IL.\todo{citation} \end{proof-sketch}
 
@@
 The proof maps every well-typed term under $\Phi'$ to a well-typed term under $\Phi$ with the same translation, and if the term has a type constructed by a tycon in $\Phi$, e.g.
-$\tc$, the mapping still has a type constructed by that tycon with the same type translation and thus we can apply the original tycon invariant.
+$\tc$, the mapping still has a type constructed by that tycon with the same type translation.
+Thus we can apply the original tycon invariant.
 The only condition on $P(\sttyidx, \iota)$ is that it be preserved under this mapping on $\sttyidx$.
 We call such propositions \emph{modular}.
-Non-modular propositions are  uninteresting because the mapping's only effect on static terms is to replace all types constructed by $\tc'$ with a unique type constructed by $\keyw{other}[m]$, for an $m$ uniquely associated with $\tc'$.
-On external terms, the mapping replaces all introductory and targeted sub-terms associated with $\tc'$ with an equivalent one that passes through the rules (ana-intro-other) and (syn-targ-other) by partially applying the intro and targeted opcon definitions to generate the term indices.
+Non-modular propositions are  uninteresting because the mapping's only effect on static terms is to replace all types constructed by $\tc'$ with a unique type constructed by $\keyw{other}[m]$, for $m$ uniquely associated with $\tc'$.
+On external terms, the mapping replaces all intro and targeted sub-terms associated with $\tc'$ with an equivalent one that passes through the rules (ana-intro-other) and (syn-targ-other) by partially applying the intro and targeted opcon definitions to generate the term indices.
 Typing, kinding, static normalization and the selective translation abstraction judgements are all preserved under this mapping, defined inductively in the supplement.
@@
 
-\newpage
+\newpage~\newpage
 % We recommend abbrvnat bibliography style.

### 2014-11-02 att-pldi15/att-pldi15.tex
More Edits
Compressed rule spacing

@@
 \def \TirNameStyle #1{\small\rulename{#1}}
-\renewcommand{\MathparLineskip}{\lineskiplimit=.6\baselineskip\lineskip=.6\baselineskip plus .2\baselineskip}
+\renewcommand{\MathparLineskip}{\lineskiplimit=.3\baselineskip\lineskip=.3\baselineskip plus .2\baselineskip}
 
@@
     \tcdef{\tc}{\psi}{\tcstruct{\stx{schema}}{\omega}} \in \Phi\\
-    \sevalNA{\stx{schema}~\sttyidx}{\sqity{\qity}}\\
+    \sevalNA{\stx{schema}(\sttyidx)}{\sqity{\qity}}\\
     \tdeabs{\Phi}{\tc}{\qity}{\memD}{\ity}{\memD'}\\
@@
     \tcdef{\tc}{\psi}{\tcstruct{\stx{schema}}{\omega}} \in \Phi\\\\
-    \sevalNA{\stx{schema}~\sttyidx}{\sqity{\qity}}\\
+    \sevalNA{\stx{schema}(\sttyidx)}{\sqity{\qity}}\\
     \tdeabs{\Phi}{\tc}{\qity}{\memD}{\ity}{\memD'}\\\\
@@
 \small\fbox{$\eanaX{e}{\st}{\iota}$}~
-\fbox{$\esynX{e}{\st}{\iota}$}
+\fbox{$\esynX{e}{\st}{\iota}$}\vspace{-3px}
 \begin{mathpar}
@@
   \sofkz{\emptyset}{\emptyset}{\Phi}{\st}{\kty}\\
-  \snormT{\st}{\st'}\\\\
+  \st \Downarrow_{\cdot; \emptyset; \Phi} \st'\\\\
   \eanaX{e}{\st'}{\iota}
@@
 \inferrule[ana-lam]{
-    \eana{\Upsilon, x \Rightarrow \st_1}{\Phi}{e}{\st_2}{\iota}\\\\
+    \eana{\Upsilon, x \Rightarrow \st_1}{\Phi}{e}{\st_2}{\iota}\\
     \sfinalrepX{\st_1}{\tau_1}
@@
     \keyw{args}(n)=\stx{args}\\\\
-  \stx{def}~\sttyidx~\stmidx~\stx{args} \Downarrow_{\es;\Upsilon;\Phi} \sqitm{\qitm}\\\\
+  \stx{def}(\sttyidx)(\stmidx)(\stx{args}) \Downarrow_{\es;\Upsilon;\Phi} \sqitm{\qitm}\\\\
   %\snorm{\stx{def}~\sttyidx~\stmidx~\stx{args}}{\emptyset}{\memG_0}{{\rightharpoonup}, \tc; \Upsilon; \Phi}{\sqitm{\itm_\text{abs}}}{\memD}{\memG}\\\\
@@
     \keyw{args}(n)=\stx{args}\\\\
-  \stx{def}~\sttyidx~\stmidx~\stx{args} \Downarrow_{(e_\text{targ}; \es);\Upsilon;\Phi} (\st, \sqitm{\qitm})\\\\
+  \stx{def}(\sttyidx)(\stmidx)(\stx{args}) \Downarrow_{(e_\text{targ}; \es);\Upsilon;\Phi} (\st, \sqitm{\qitm})\\\\
   %\snorm{\stx{def}~\sttyidx~\stmidx~\stx{args}}{\emptyset}{\memG_0}{{\rightharpoonup}, \tc; \Upsilon; \Phi}{\sqitm{\itm_\text{abs}}}{\memD}{\memG}\\\\
@@
     \keyw{args}(n)=\stx{args}\\
-    \stx{def}~\stx{args} \Downarrow_{\es; \Upsilon; \Phi} \sqitm{\qitm}\\\\
+    \stx{def}(\stx{args}) \Downarrow_{\es; \Upsilon; \Phi} \sqitm{\qitm}\\\\
     \edeabs{\keyw{other}[m]}{\es; \Upsilon;\Phi}{\qitm}{\emptyset}{\emptyset}{\itm_\text{abs}}{\memD}{\memG}\\\\
@@
     \keyw{args}(n)=\stx{args}\\
-  \stx{def}~\stx{args} \Downarrow_{(e_\text{targ}; \es);\Upsilon;\Phi} (\st, \sqitm{\qitm})\\\\
+  \stx{def}(\stx{args}) \Downarrow_{(e_\text{targ}; \es);\Upsilon;\Phi} (\st, \sqitm{\qitm})\\\\
   %\snorm{\stx{def}~\sttyidx~\stmidx~\stx{args}}{\emptyset}{\memG_0}{{\rightharpoonup}, \tc; \Upsilon; \Phi}{\sqitm{\itm_\text{abs}}}{\memD}{\memG}\\\\
@@
      \esynX{\etarg{\opname{op}}{\stx{def}}{e_\text{targ}}{\es}}{\st}{[\delta][\gamma]\iota_\text{abs}}
-}\vspace{-5px}
+}\vspace{-8px}
 \end{mathpar}
@@
 
-The remaining premises of (syn-targ) are analagous to the corresponding premises in (ana-intro), with the only difference being that we check the selectively abstract term translation against the selectively abstracted type translation of the synthesized type, but the delegated tycon is that of the type synthesized by the target argument.
+The remaining premises of (syn-targ) are analagous to the corresponding premises in (ana-intro), with the only difference being that we check the selectively abstracted.
+term translation against the selectively abstracted type translation of the synthesized type, but the delegated tycon is that of the type synthesized by the target argument.
 
@@
 
-\newpage~\newpage
+\newpage
 % We recommend abbrvnat bibliography style.

### 2014-11-02 att-pldi15/att-pldi15.tex
Edits to related work

@@
         \slam{\kargs}{\svar{args}}{\\
-            \quad \keyw{letpair}~(\svar{arg1}, \svar{arg2}) = \svar{arity2}~\svar{args}\\
-            \quad \keyw{letpair}~(\_, \svar{tr1}) = \svar{syn}~\svar{arg1}\\
-            \quad \keyw{letpair}~(\svar{ty2}, \svar{tr2}) = \svar{syn}~\svar{arg2}\\
+            \quad \keyw{letpair}~(\svar{arg1}, \svar{arg2}) = \svar{arity2}~\svar{args}~\keyw{in}\\
+            \quad \keyw{letpair}~(\_, \svar{tr1}) = \svar{syn}~\svar{arg1}~\keyw{in}~ \keyw{letpair}~(\svar{ty2}, \svar{tr2}) = \svar{syn}~\svar{arg2}\\
             \quad \stycase{\tcvar{rstr}}{\svar{ty2}}{\svar{tyidx2}}{\\
@@
 
-The reason why the rule (syn-targ-other) is never a problem in proving a tycon invariant turns out to be the same reason extending the tycon context conserves all tycon invariants.
+The reason why the rule (syn-targ-other) is never a problem in proving a tycon invariant, i.e.
+\emph{translation independence} of tycons, turns out to be the same reason extending the tycon context conserves all tycon invariants.
 A newly introduced tycon defining a targeted operator that synthesizes a regular string type, e.g.
@@
 For example, a refinement of  integers might distinguish negative integers.
-Proposals for ``pluggable type systems'' describe such type refinement systems \cite{Brac04a,Andreae:2006:FIP:1167473.1167479}.
+Proposals for ``pluggable type systems'' describe composing such type refinement systems \cite{Brac04a,Andreae:2006:FIP:1167473.1167479}.
 Refinements of abstract types can be used to hold representations abstract.
@@
 Proof assistants and logical frameworks can be used to mechanize dialect specification and metatheory, but they similarly require a complete specification.
-Chlipala has suggested using  proof automation that covers a variety of language features to decrease the burden when a  dialect is formed, but this is a fundamentally heuristic approach \cite{Chlipala10}\todo{wrong citation}.
+Chlipala has suggested using  proof automation that covers a variety of language features to decrease the burden when forking a  dialect, but this is a fundamentally heuristic approach \cite{Chlipala10}\todo{wrong citation}.
 Our approach provides a conservativity theorem.
-No proof terms are needed to guarantee modular reasoning:   mistakes can only weaken the invariants at a single tycon.
+No proof terms are needed to guarantee modular reasoning:   mistakes in a tycon definition can only weaken invariants at that tycon.
 %This facilitates exploratory language design and suggests educational applications.
-A promising avenue for future work beyond the scope of this paper is to introduce optional proof mechanization into the SL (by making it dependently kinded), and to mechanize the metatheory of @$\lambda$ itself.%Every combination of fragments is a new dialect, and must be reasoned about monolithically.
+A promising avenue for future work is to introduce optional proof mechanization into the SL (by making it dependently kinded), and to mechanize the metatheory of @$\lambda$ itself.%Every combination of fragments is a new dialect, and must be reasoned about monolithically.
 %Efforts to reason modularly areeed, several problems can come up when fragments are combined, so .
@@
 
-Type abstraction is, encouragingly, also the technique underlying modular reasoning in ML-like languages.
-Indeed, proofs of tycon invariants rely on the same abstraction theorem.
+Type abstraction is, encouragingly, also the technique underlying modular reasoning in ML-like languages\todo{citation}.
+Indeed, proofs of tycon invariants can rely on the same abstraction theorem.
 Our work has some technical similarities to work on syntactic type abstraction \cite{journals/toplas/GrossmanMZ00} and on translating modules to System F \cite{conf/tldi/RossbergRD10}.

### 2014-11-02 att-pldi15/att-pldi15-supplement.tex
Updated related work

@@
 \section{Internal Language}
-\todo{copy from one of the 312 HWs}
+We assume the statics of the IL are specified in the standard way by judgements for  type formation {$\iDelta \vdash \tau$}, typing context formation { $\iDelta \vdash \iGamma$} and type assignment {$\iDelta~\iGamma \vdash \iota : \tau\moutput$}.
+%The typing and type formation contexts obey standard structural properties (i.e. weakening, exchange and contraction). 
+The internal dynamics are specified as a structural operational semantics with a stepping judgement {\small $\iota \mapsto \iota\moutput$} and a value judgement {$\iota~\mathtt{val}$}.
+The multi-step judgement $\iota \mapsto^{*} \iota\moutput$ is the reflexive, transitive closure of the stepping judgement and the evaluation judgement $\iota \Downarrow \iota'$ is defined iff $\iota \mapsto^{*} \iota'$ and $\iota'~\mathtt{val}$.
+Both the static and dynamic semantics of the IL can be found in any standard textbook covering typed lambda calculi (we directly follow \cite{pfpl}), so we assume familiarity and give the lemmas in this section without proof.
 
-\subsection{Substitutions}
+We use {$\mathcal{L}\{{\rightharpoonup}\,{\forall}\,{\mu}\,{\iunit}\,{\times}\,{+}\}$}, the syntax for which is shown in Figure \ref{syntax-IL}, as representative of any intermediate language for a typed functional language.
+In fact, our intention is not to prescribe a particular choice of IL, so we will here only review the key metatheoretic properties that the IL must possess.
+Each choice of IL is technically a distinct dialect of @$\lambda$, but for the broad class of ILs that enjoy these properties, the metatheory in the remainder of the supplement should follow without trouble.
 
-\subsection{Abstraction Theorem}
-\todo{hm...}
+\begin{lemma}[Internal Type Assignment]
+If $\Delta \vdash \Gamma$ and $\Delta~\Gamma \vdash \iota : \tau$ then $\Delta \vdash \tau$.
+\end{lemma}
+
+Internal type safety follows the standard methodology of proving preservation and progress lemmas.
+
+\begin{lemma}[Internal Progress]
+If $\emptyset~\emptyset \vdash \iota : \tau$ then either $\iota~\texttt{val}$ or $\iota \mapsto \iota'$.
+\end{theorem}
+\begin{lemma}[Internal Preservation]
+If $\emptyset~\emptyset \vdash \iota : \tau$ and $\iota \mapsto \iota'$ then $\emptyset~\emptyset \vdash \iota' : \tau$. 
+\end{lemma}
+
+We assume substitution and simultaneous $n$-ary substitutions, $\delta$ and $\gamma$, have the standard semantics, and that typing and type formation contexts, $\Delta$ and $\Gamma$, obey standard properties of weakening, exchange and contraction.
+We assume that the names of variables and type variables are unimportant, so that $\alpha$-equivalent terms, types and contexts can be identified implicitly throughout this work.
+In particular, we need the definitions of substitution validity shown in Figures \ref{fig:internal-type-substitution-validity} and \ref{fig:internal-term-substitution-validity} and the following lemmas.
+
+\begin{figure}[t]
+\small
+\hspace{-5px}\input{syntax-table-IL}
+\caption{Syntax of {$\mathcal{L}\{{\rightharpoonup}\,{\forall}\,{\mu}\,{\iunit}\,{\times}\,{+}\}$}, our internal language (IL).
+Metavariable $x$ ranges over term variables and $\alpha$ and $t$ both range over type variables.}
+\label{syntax-IL}
+\end{figure}
+
+\begin{figure}
+\small\fbox{$\Delta \vdash \delta : \Delta$}~~$\delta ::=  \emptyset ~|~ \delta, \tau/\alpha$
+
+\begin{mathpar}
+\inferrule[tysub-emp]{ }{\Delta \vdash \emptyset : \emptyset}
+
+\inferrule[tysub-ext]{
+    \Delta \vdash \delta : \Delta'\\
+    \Delta \vdash \tau
+}{
+    \Delta \vdash \delta, \tau/\alpha : \Delta', \alpha
+}
+\end{mathpar}
+\caption{Internal Type Substitution Validity}
+\label{fig:internal-type-substitution-validity}
+\end{figure}
+\begin{figure}
+\small\fbox{$\Delta~\Gamma \vdash \gamma : \Gamma$}~~$\gamma ::= \emptyset ~|~ \gamma, \iota/x$
+\begin{mathpar}
+\inferrule[tmsub-emp]{ }{\Delta~\Gamma \vdash \emptyset : \emptyset}
+
+\inferrule[tmsub-ext]{
+    \Delta~\Gamma \vdash \gamma : \Gamma'\\
+    \Delta \vdash \tau\\
+    \Delta~\Gamma \vdash \iota : \tau
+}{
+    \Delta~\Gamma \vdash \gamma, \iota/x : \Gamma', x : \tau
+}
+\end{mathpar}
+\caption{Internal Term Substitution Validity}
+\label{fig:internal-term-substitution-validity}
+\end{figure}
+\begin{lemma}[Internal Type Substitution on Types] If $\Delta \vdash \delta : \Delta'$ and $\Delta\Delta' \vdash \tau$ then $\Delta \vdash [\delta]\tau$.\end{lemma}
+
+\begin{lemma}[Internal Type Substitutions on Typing Contexts] If $\Delta \vdash \delta : \Delta'$ and $\Delta\Delta' \vdash \Gamma$ then $\Delta \vdash [\delta]\Gamma$.
+\end{lemma}
+
+\begin{lemma}[Internal Type Substitutions on Terms] If $\Delta \vdash \delta : \Delta'$ and $\Delta\Delta' \vdash \tau$ and $\Delta\Delta' \vdash \Gamma$ and $\Delta\Delta'~\Gamma \vdash \iota : \tau$ then $\Delta~[\delta]\Gamma \vdash [\delta]\iota : [\delta]\tau$.
+\end{lemma}
+
+\begin{lemma}[Internal Term Substitutions] If $\Delta \vdash \Gamma$ and $\Delta \vdash \Gamma'$ and $\Delta~\Gamma \vdash \gamma : \Gamma'$ and $\Delta~\Gamma\Gamma' \vdash \iota : \tau$ then $\Delta~\Gamma \vdash [\gamma]\iota : \tau$.
+\end{lemma}
 
 \section{Tycon Contexts}
+\begin{figure}\fbox{$\vdash \Phi$}\vspace{-25px}
+\begin{mathpar}
+\small
+\inferrule[tcc-emp]{ }{\vdash \cdot}
+
+\inferrule[tcc-ext]{
+    \vdash \Phi\\
+    \tc \notin \text{dom}(\Phi)\\
+    \keq{\emptyset}{\ktyidx}\\
+    \sofkz{\emptyset}{\emptyset}{\Phi}{\stx{schema}}{\ktyidx \rightarrow \kity}\\\\
+    \vdash_{\Phi,  \tcdef{\tc}{\tcsig{\ktyidx}{\chi}}{\tcstruct{\stx{schema}}{\omega}}} \omega \sim \tcsig{\ktyidx}{\chi}
+}{
+    \vdash \Phi, \tcdef{\tc}{\tcsig{\ktyidx}{\chi}}{\tcstruct{\stx{schema}}{\omega}}
+}
+\end{mathpar}
+\caption{Tycon context well-definedness.}
+\label{fig:tycon-ctxs}
+\end{figure}
+\begin{figure}[t]
+\small\fbox{$\vdash_\Phi \omega \sim \psi$}
+\begin{mathpar}\small
+\inferrule[ocstruct-intro]{
+    \introsig{\klitidx} \in \chi\\
+    \emptyset \vdash \klitidx\\\\
+    \sofkz{\emptyset}{\emptyset}{\Phi}{\stx{def}}{\ktyidx \rightarrow \klitidx \rightarrow \kargs \rightarrow \kitm}
+}{
+    \vdash_\Phi \tcstructn{\stx{def}} \sim \tcsig{\ktyidx}{\chi}
+}
+
+\inferrule[ocstruct-targ]{
+    \vdash_\Phi \omega \sim \tcsig{\ktyidx}{\chi}\\
+    \opname{op}\notin\text{dom}(\chi)\\
+    \emptyset \vdash \klitidx\\\\
+    \sofkn{\emptyset}{\emptyset}{\Phi}{0}{\stx{def}}{\ktyidx \rightarrow \klitidx \rightarrow \kargs \rightarrow (\kty \times \kitm)}
+}{
+    \vdash_\Phi \tcstructc{\omega}{op}{\stx{def}} \sim \tcsig{\ktyidx}{\chi, \opsig{\opname{op}}{\klitidx}}
+}
+\end{mathpar}
+\caption{Checking opcon structures against tycon signatures.}
+\label{ocstruct}
+\end{figure}
+
 \subsection{Tycon Context Well-Definedness}

### 2014-11-02 att-pldi15/att-pldi15.tex
Updated related work

@@
 For example, a module defining sets in ML can hold the representation of sets abstract, ensuring that any invariants maintained by the functions in the module (e.g.
-uniqueness, if using a list representation) will hold no matter which other modules are separately in use by a client.
-Other langauges offer similar facilities, sometimes called \emph{abstract data types}.\todo{citation?} %Mechanisms that can help decrease the complexity of an embedding without violating abstraction barriers are thus valuable, and we will lead into our work in Sec.
+uniqueness, if using a list representation) will hold no matter which other modules are separately in use by a client \cite{harper1997programming}.
+Other langauges offer similar facilities, often grouped under the moniker of \emph{abstract data types} \cite{liskov1974programming}.
+%Mechanisms that can help decrease the complexity of an embedding without violating abstraction barriers are thus valuable, and we will lead into our work in Sec.
 \ref{prior-work} by summarizing them.
@@
 The first is once again to justify the practicality of our approach: local type inference is increasingly being used in modern languages (e.g.
-Scala\todo{citation}) because it eliminates the need for type annotations in many situations while remaining decidable in more situations than whole-function type inference and providing what are widely perceived to be higher quality error messages\todo{citation from Dunfield}.
-Secondly, it will give us a clean way to reuse the generalized introductory form, $\eintro{\st}{\es}$, and its associated desugarings, at many types.
-For example, regular string types will be able to use standard string literal syntax.
+Scala \cite{OdeZenZen01}) because it eliminates the need for type annotations in many situations while remaining decidable in more situations than whole-function type inference and providing what are widely perceived to be higher quality error messages \cite{journals/jfp/JonesVWS07}.
+Secondly, it  gives us a clean way to reuse the generalized introductory form, $\eintro{\st}{\es}$, and its associated desugarings, at many types \cite{TSLs}.
+For example, regular string types can use standard string literal syntax.
 
@@
 We define these by the judgement $\keq{\kDelta}{\kappa}$ (see supplement).
-Equality kinds are similar to equality types as found in Standard ML\todo{citation}. The main implication of this choice is that type indices cannot contain static functions.
+Equality kinds are similar to equality types as found in Standard ML \cite{Tofte:89:TheDefinitionOfStandardML}.
+The main implication of this choice is that type indices cannot contain static functions.
 
@@
 Type safety for the IL \cite{pfpl} then implies that well-typed external terms cannot go wrong.
-To prove this, we must actually prove a stronger theorem, \emph{type-preserving translation}, analagous to type-preserving compilation in the typed compilation literature \cite{tarditi+:til-OLD}\todo{right citation for TIL?}:% produced as the translation of the external type.
+To prove this, we must actually prove a stronger theorem, \emph{type-preserving translation}, analagous to type-preserving compilation in the typed compilation literature \cite{tarditi+:til-OLD}:% produced as the translation of the external type.
 Note that we need only to state our top-level theorems for the analysis judgement  because of the subsumption rule.
@@
 The interesting cases are (ana-intro), (ana-intro-other), (syn-trans) and (syn-trans-other); the latter two arise via subsumption.
-The result follows directly from the final premise of each rule, combined with lemmas that state that if $\memD \leadsto \delta : \Delta_\text{abs}$ and $\memG \leadsto \gamma : \Gamma_\text{abs}$ then $\vdash \delta : \Delta_\text{abs}$ and $\Delta_\text{abs}~\Gamma \vdash \gamma : \Gamma_\text{abs}$, i.e.
-that all the variables in $\Delta_\text{abs}$ and $\Gamma_\text{abs}$ have well-formed/well-typed substitutions in $\delta$ and $\gamma$,  and thus that applying them in the conclusions of the rules gives a well-typed term.
+The result follows directly from the final premise of each rule, combined with lemmas that state that if $\memD \leadsto \delta : \Delta_\text{abs}$ and $\memG \leadsto \gamma : \Gamma_\text{abs}$ then $\emptyset \vdash \delta : \Delta_\text{abs}$ and $\Delta_\text{abs}~\Gamma \vdash \gamma : \Gamma_\text{abs}$, i.e.
+that all variables in $\Delta_\text{abs}$ and $\Gamma_\text{abs}$ have well-formed/well-typed substitutions in $\delta$ and $\gamma$,  and so applying them in the conclusions of the rules gives a well-typed term.
 \end{proof-sketch}
@@
 \begin{proof-sketch} The proof is not unusually difficult because we have fixed the tycon context $\Phi_\text{rstr}$, so we can simply treat the calculus like a type-directed compiler for a calculus with only two tycons, $\rightharpoonup$ and $\tcvar{rstr}$.
-Such a calculus and compiler specification was given by Fulton et al.
-\cite{sanitation-psp14}, so we must simply show that the opcon definitions in $\tcvar{rstr}$ adequately capture these specifications using standard program correctness techniques for the SL, a simply-typed functional language.
+Such a calculus and compiler specification was given in \cite{sanitation-psp14}, so we must simply show that the opcon definitions in $\tcvar{rstr}$ adequately capture these specifications using standard techniques for the SL, a simply-typed functional language \cite{conf/pldi/Chlipala07}.
 The only ``twist'' is that the rule (syn-targ-other) can synthesize a regular string type.
-But if so, the abstracted term translation will be checked against $\tau_\text{abs}=\alpha$, so only strings that could otherwise arise can result, by the standard abstraction theorem for the IL.\todo{citation} \end{proof-sketch}
+But if so, the abstracted translation must have type $\tau_\text{abs}=\alpha$.
+Thus, invariants cannot be violated by relational parametricity in the IL (i.e.
+a ``free theorem'') \cite{WadlerThms}.
+\end{proof-sketch}
 
@@
 As discussed throughout the paper, our treatment of concrete syntax defers to recent work on \emph{type-specific languages} \cite{TSLs}.
-As in our work, TSLs determine the meaning of introductory terms based on the type the term is analyzed against, but TSLs only support syntactic elaboration to a fixed semantics.
+%As in our work, TSLs determine the meaning of introductory terms based on the type the term is analyzed against, but TSLs only support syntactic elaboration to a fixed semantics.
 
@@
 %\paragraph{Optimization}
-When a simple embedding that preserves the static semantics exists, but a different  embedding is needed to preserve a desired cost semantics, term rewriting  can also be used to perform ``optimizations'', thus achieving an isomorphic embedding.
-Care is needed when the optimized value has a different type, to ensure that the invariants of the unoptimized term are preserved.
+When a simple embedding that preserves the static semantics exists, but a different  embedding is needed to preserve a desired cost semantics, term rewriting  can also be used to perform ``optimizations'', achieving an isomorphic embedding.
+Care is needed when this changes the type of a term to ensure that invariants are preserved.
 Type abstraction has been used for this purpose in work on \emph{lightweight modular staging} (LMS) \cite{Rompf:2012:LMS}.
@@
 These either do not support forming languages from separately defined fragments or provide few clear modular reasoning principles that make it possible to reason separately about them.
+\todo{read and describe amal ahmed's recent paper}
 
 Proof assistants and logical frameworks can be used to mechanize dialect specification and metatheory, but they similarly require a complete specification.
-Chlipala has suggested using  proof automation that covers a variety of language features to decrease the burden when forking a  dialect, but this is a fundamentally heuristic approach \cite{Chlipala10}\todo{wrong citation}.
+Chlipala has suggested using  proof automation that covers a variety of language features to decrease the burden when forking a  dialect, but this is a fundamentally heuristic approach \cite{Chlipala10}.
 Our approach provides a conservativity theorem.
@@
 
-Type abstraction is, encouragingly, also the technique underlying modular reasoning in ML-like languages\todo{citation}.
-Indeed, proofs of tycon invariants can rely on the same abstraction theorem.
-Our work has some technical similarities to work on syntactic type abstraction \cite{journals/toplas/GrossmanMZ00} and on translating modules to System F \cite{conf/tldi/RossbergRD10}.
+Type abstraction is, encouragingly, also the technique underlying modular reasoning in ML-like languages.
+Indeed, proofs of tycon invariants can rely on the same parametricity theorem \cite{WadlerThms}.
+Our work has technical similarities to work on  translating modules to System $\mathbf{F}_\omega$ \cite{conf/tldi/RossbergRD10}.
 Unlike in module systems, type translations (analagous to the choice of representation for an abstract type) can be \emph{computed} based on a type index.
 Moreover, there can be arbitrarily many operators because they arise by providing a term index to an opcon.
-Their static and dynamic semantics can be complex because a static function computes the type and translation.
+Their semantics can be complex because a static function computes the type and translation.
 In contrast, modules can only specify a fixed number of operations, and each must have function type.
 Note that although we did not specify quantification over external types here for simplicity, we conjecture that it is complementary (and thus @$\lambda$ could serve as the core of a language with a ML-style module system).
-Another avenue for future work is tycon functors, which abstract over tycons with the same tycon signature.
+Another direction is \emph{tycon functors}, which abstract over tycons with the same  signature.
 % This permits us tocode targeting a separate IL, and permit operations whose types cannot simply be written as functions.

### 2014-11-03 att-pldi15/att-pldi15.tex
Compressed a bit more

@@
 \renewcommand\floatpagefraction{0.85}
+
+\AtBeginDocument{%
+ \abovedisplayskip=5pt
+ \abovedisplayshortskip=4pt
+ \belowdisplayskip=4pt
+ \belowdisplayshortskip=4pt
+}
 
@@
 Instead, the approach we take relies on type abstraction in the internal language.
-As a result, we are able to borrow the same results that underly modular reasoning in simply-typed languages like ML to reason modularly about typed language fragments.
-We conclude by discussing  related and future work in Sec.
+As a result, we are able to use the same parametricity results that underly modular reasoning in simply-typed languages like ML to reason modularly about typed language fragments.
+We conclude with  related and future work in Sec.
 \ref{prior-work}.
@@
 \section{Overview of @$\lambda$}\label{atlam}\label{overview}
+\begin{figure}[t]
+\small
+\hspace{-5px}\input{syntax-table-IL}
+\caption{Syntax of {$\mathcal{L}\{{\rightharpoonup}\,{\forall}\,{\mu}\,{\iunit}\,{\times}\,{+}\}$}, our internal language (IL).
+Metavariable $x$ ranges over term variables and $\alpha$ and $t$ both range over type variables.}
+\label{syntax-IL}
+\vspace{-10px}
+\end{figure}
 \paragraph{External Language}
@@
 The static and dynamic semantics are given simultaneously as a \emph{bidirectionally typed translation semantics}, i.e.
-the key judgements take the form:  \[\esynX{e}{\st\moutput}{\iota\moutput} ~~~~~\text{and}~~~~~ \eanaX{e}{\st}{\iota\moutput}\]
+the key judgements take the form:\[\esynX{e}{\st\moutput}{\iota\moutput} ~~~~~\text{and}~~~~~ \eanaX{e}{\st}{\iota\moutput}\]
 \noindent
-
 These are pronounced ``$e$ (synthesizes / analyzes against) type $\sigma$ and has  translation $\iota$ under typing context $\Upsilon$ and tycon context $\Phi$''.
@@
 External typing contexts $\Upsilon$ map variables to types, so we also need the judgement $\vdash_\Phi \Upsilon \leadsto \Gamma$, where $\Gamma$ maps variables to the corresponding type translation.
-In examples, we omit leading $\emptyset$, used as the base case for finite mappings, and $\cdot$, used as the base case for finite sequences, for concision.
 
@@
 % \end{figure*}
-\begin{figure}[t]
-\small
-\hspace{-5px}\input{syntax-table-IL}
-\caption{Syntax of {$\mathcal{L}\{{\rightharpoonup}\,{\forall}\,{\mu}\,{\iunit}\,{\times}\,{+}\}$}, our internal language (IL).
-Metavariable $x$ ranges over term variables and $\alpha$ and $t$ both range over type variables.}
-\label{syntax-IL}
-\vspace{-10px}
-\end{figure}
 
@@
 \hspace{-5px}\input{syntax-table-EL}
-\caption{Syntax of the external language (EL).}\label{syntax-EL}
+\caption{Syntax of the external language (EL).}
+\label{syntax-EL}\vspace{-5px}
 \end{figure}
@@
 The kinding judgement takes the form $\sofkX{\st}{\kappa\moutput}$, where $\kDelta$ and $\kGamma$ are analagous to $\iDelta$ and $\iGamma$ and analagous kind and kinding context formation judgements $\kDelta \vdash \kappa$ and $\kDelta \vdash \kGamma$ are  defined.
+All such contexts in @$\lambda$ are identified up to exchange and contraction and obey weakening \cite{pfpl}.
 The natural number $n$ is used as a technical device in our semantics to ensure that the forms shown as being indexed by $n$ in the syntax only arise in a controlled manner internally to prevent ``out of bounds'' issues, as we will discuss; they would have no corresponding concrete syntax so $n$ can be assumed $0$ in user-defined terms.
@@
 \textbf{opcon sigs}& \chi & ::= & \introsig{\kappa} ~|~ \opsigS{\theta}{op}{\kappa}\\
-\end{array}$
+\end{array}\vspace{-3px}$
 \caption{Syntax of tycons. Metavariables $\tc$ and $\opname{op}$ range over user-defined tycon and opcon names, respectively, and $m$ ranges over natural numbers.}
-\label{syntax-TC}\vspace{-5px}
+\label{syntax-TC}\vspace{-8px}
 \end{figure}
@@
 \caption{Kinding rules for types, which take the form $\sty{c}{\sttyidx}$ where $c$ is a tycon and $\sttyidx$ is the type index.}
-\label{fig:types}\vspace{-5px}
+\label{fig:types}\vspace{-8px}
 \end{figure*}
@@
 The type indices are  here written using standard concrete syntax for concision; recent work has specified how to define type-specific (or here, kind-specific) syntax like this composably in libraries \cite{TSLs}.
-We define the tycon context containing only the definition of $\tcvar{rstr}$, $\Phi_\text{rstr} := \tcdef{\tcvar{rstr}}{\tcsig{\krx}{\chi_\text{rstr}}}{\theta_\text{rstr}}$.
+We define the tycon context containing only the definition of $\tcvar{rstr}$, $\Phi_\text{rstr} := \tcdef{\tcvar{rstr}}{\tcsig{\krx}{\chi_\text{rstr}}}{\theta_\text{rstr}}$.\footnote{In examples, we omit leading $\emptyset$, used as the base case for finite mappings, and $\cdot$, used as the base case for finite sequences, for concision.
+}
 
@@
 \subsection{Static Type Case Analysis}
+\begin{figure}\hfill \fbox{$\vdash \Phi$}\vspace{-25px}\begin{mathpar}
+\small
+\inferrule[tcc-ext]{
+    \vdash \Phi\\
+    \tc \notin \text{dom}(\Phi)\\
+    \keq{\emptyset}{\ktyidx}\\
+    \sofkz{\emptyset}{\emptyset}{\Phi}{\stx{schema}}{\ktyidx \rightarrow \kity}\\\\
+    \vdash_{\Phi,  \tcdef{\tc}{\tcsig{\ktyidx}{\chi}}{\tcstruct{\stx{schema}}{\omega}}} \omega \sim \tcsig{\ktyidx}{\chi}
+}{
+    \vdash \Phi, \tcdef{\tc}{\tcsig{\ktyidx}{\chi}}{\tcstruct{\stx{schema}}{\omega}}
+}\vspace{-10px}
+\end{mathpar}
+\caption{Tycon context well-definedness. We omit the trivial case for when $\Phi=\cdot$ for concision.}
+\label{fig:tycon-ctxs}
+\vspace{-8px}
+\end{figure}
+
 Types in @$\lambda$ can be thought of as arising from a distinguished ``open datatype'' defined by the tycon context \cite{conf/ppdp/LohH06}.
@@
 The dynamics (see supplement) are straightforwardly consistent with the intuition above.% We will see an example of its use in Sec. \ref{sec:targops}.
-\begin{figure}\hfill \fbox{$\vdash \Phi$}\vspace{-25px}\begin{mathpar}
-\small
-\inferrule[tcc-ext]{
-    \vdash \Phi\\
-    \tc \notin \text{dom}(\Phi)\\
-    \keq{\emptyset}{\ktyidx}\\
-    \sofkz{\emptyset}{\emptyset}{\Phi}{\stx{schema}}{\ktyidx \rightarrow \kity}\\\\
-    \vdash_{\Phi,  \tcdef{\tc}{\tcsig{\ktyidx}{\chi}}{\tcstruct{\stx{schema}}{\omega}}} \omega \sim \tcsig{\ktyidx}{\chi}
-}{
-    \vdash \Phi, \tcdef{\tc}{\tcsig{\ktyidx}{\chi}}{\tcstruct{\stx{schema}}{\omega}}
-}\vspace{-8px}
-\end{mathpar}
-\caption{Tycon context well-definedness. We omit the trivial case for when $\Phi=\cdot$ for concision.}
-\label{fig:tycon-ctxs}
-\vspace{-5px}
-\end{figure}
 \subsection{Tycon Context Well-Definedness}
@@
 \end{mathpar}
-The unquote form is eliminated during normalization, while references to the translation of a type are retained in values of kind $\kity$. The key rules in the dynamics are: 
+The unquote form is eliminated during normalization, while references to type translations are retained in values:
 \begin{mathpar}
@@
 \end{mathpar}
-
-These are used in the schema for $\tcvar{lprod}$, which generates nested binary product types by recursing over the type index and referring to the translations of the types therein (this is not the only viable choice, e.g.
+These forms are needed in the translation schema for $\tcvar{lprod}$, which generates nested binary product types by recursing over the type index and referring to the translations of the types therein (this is not the only viable choice, e.g.
 we could also have used a list).
@@
 \end{array}\]
-Evaluating this translation schema with the index of $\stx{paper}$, for example, produces the value $\stx{paper/trans} := \sqity{\qity_\text{paper/trans}}$ where $\small\qity_\text{paper/trans} := \srep{\stx{title}}\times(\srep{\stx{conf}}\times \iunit)$.
+Applying this translation schema to the index of $\stx{paper}$, for example, produces the value $\stx{paper/trans} := \sqity{\qity_\text{paper/trans}}$ where $\small\qity_\text{paper/trans} := \srep{\stx{title}}\times(\srep{\stx{conf}}\times \iunit)$.
 Note that we do not include logic to optimize away the trailing unit type for simplicity (and to again emphasize that many translations are possible for any given type).%For simplicity, and to make the point that the choice of representation has only tycon-local implications, our translation schema does not optimize away
@@
 
-Variables and functions behave in the standard manner given our definitions of types and type translations (used to generate ascriptions in the IL, which is intrinsically rather than bidirectionally typed).
+Variables and functions behave in the standard manner given our definitions of types and type translations (used to generate ascriptions in the IL).
 We use Plotkin's fixpoint operator for general recursion (cf.
-\cite{pfpl}), and define both lambdas and fixpoints only in analytic positions for simplicity.
+\cite{pfpl}), and define both lambdas and fixpoints only analytically for simplicity.
 
 \subsection{Generalized Introductory Operations}\label{sec:introop}
-The translation of the generalized introductory form, $\eintro{\sttmidx}{\es}$, is determined by the tycon of the type it is being analyzed against as a function of the type's index, the \emph{term index}, $\sttmidx$, and the \emph{argument list}, $\es$.
+The translation of the \emph{generalized intro operation}, written $\small\eintro{\sttmidx}{\es}$, is determined by the tycon of the type it is being analyzed against as a function of the type's index, the \emph{term index}, $\sttmidx$, and the \emph{argument list}, $\es$.
 
@@
 Next, it checks the string provided as the term index against the regular expression given as the type index using $\svar{rmatch} :: \krx \rightarrow \keyw{Str} \rightarrow \kunit$, which we assume is defined in the usual way and again raises an error on failure.
-Finally, a {translation} corresponding to the term index is generated via  helper function $\svar{str\_of\_Str} :: \keyw{Str} \rightarrow \kitm$.% and $\svar{nat\_of\_Nat} :: \keyw{Nat} \rightarrow \kitm$ to generate  translational internal terms corresponding to the provided static terms and $\svar{len} :: \keyw{Str} \rightarrow \keyw{Nat}$ to statically compute the length of the string.
+Finally, the \emph{translational internal string} corresponding to the static string provided as the term index is generated via the helper function $\svar{str\_of\_Str} :: \keyw{Str} \rightarrow \kitm$.% and $\svar{nat\_of\_Nat} :: \keyw{Nat} \rightarrow \kitm$ to generate  translational internal terms corresponding to the provided static terms and $\svar{len} :: \keyw{Str} \rightarrow \keyw{Nat}$ to statically compute the length of the string.
 
@@
 Before giving the rules, let us motivate the mechanism with the intro opcon definition for $\tcvar{lprod}$.
-We have $\omega_\text{lprod} := \tcstructn{\stx{lprod/intro}}; \omega_\text{lprod/targops}$ where $\stx{lprod/intro} :=$\vspace{-5px}
+We have $\omega_\text{lprod} := \tcstructn{\stx{lprod/intro}}; \omega_\text{lprod/targops}$ where $\stx{lprod/intro} :=$
 \[\small
@@
 \quad\quad\quad \sqitm{(\quq{\svar{rowtr}}, \quq{\svar{r}})
-}
-\end{array}
+}\end{array}
 \]
+
 The first line checks that the type provided is inhabited, in this case by checking that there are no duplicate labels via the helper function $\svar{uniqmap} :: \klist{\klbl \times \kty}\rightarrow \kunit$, raising an error if there are.
@@
 We chose not to formalize these for simplicity, and to demonstrate this general technique.
-%An analagous technique could be used to implement record types by requiring that the index be sorted.%, leaving indices where the labels did not appear sorted uninhabited.
+%An analagous technique could be used to implement record types by requiring that the index be sorted (see supplement for additional discussion).%, leaving indices where the labels did not appear sorted uninhabited.
 In both cases, we could provide a static helper function of kind $\klist{\klbl\times\kty}\rightarrow \kty$ that checked well-formedness immediately before constructing the requested type.
@@
 
-The argument interface list is is then simply a static value of kind $\kargs$ of length $n$ where the $i$th entry is $(\slam{\kty}{\svar{ty}}{\sana{i}{\svar{ty}}}, \slam{\kunit}{\_}{\ssyn{i}})$.
+The \emph{argument interface list} is then simply a static value of kind $\kargs$ of length $n$ where the $i$th entry is the argument interface $(\slam{\kty}{\svar{ty}}{\sana{i}{\svar{ty}}}, \slam{\kunit}{\_}{\ssyn{i}})$.
 It is generated by the judgement $\keyw{args}(n)=\stx{args}$, where $n$ is the length of the argument list, written $|\es|=n$.
@@
 \end{mathpar}
-
 The following lemma characterizes thus characterizes the argument list interface generation judgement:
@@
 The rule (ocstruct-intro) ruled out writing either of these forms explicitly in an opcon definition by checking against the bound $n=0$.
-This is to prevent out-of-bounds errors -- tycon providers do not write these forms directly, only accessing them via the argument interface list, guaranteed to have the correct length.
-This is  safe because we can weaken the upper bound:
+This is to prevent out-of-bounds errors: tycon providers do not write these forms directly, only accessing them via the argument interface list, which is guaranteed to have the correct length.
+This is  safe because $n$ is only an upper bound:
 \begin{lemma}
@@
 These forms serve as the link between the dynamics of the static language and the statics of the external language.
-For $\sana{n}{\st}$, after stepping the type $\st$ to a value and propagating errors, the argument environment, which stores the arguments themselves and the typing and tycon contexts, $\es; \Upsilon; \Phi$, is consulted to retrieve the $n$th argument and analyze it against $\st$.
-If this succeeds, the translational internal term $\sqitm{\anatrans{n}{\st}}$ is generated to refer to it.
+For $\sana{n}{\st}$, after normalizing $\st$, the argument environment, which contains the arguments themselves and the typing and tycon contexts, $\argEnv ::= \es; \Upsilon; \Phi$, is consulted to retrieve the $n$th argument and analyze it against $\st$.
+If this succeeds, the translational internal term $\sqitm{\anatrans{n}{\st}}$ is generated:
 \begin{mathpar}\small
@@
 
-The kinding rules also prevent these translational internal term forms  from being well-kinded when $n = 0$.
-Like the translational internal type form $\srep{\st}$, these  are retained in values of kind $\kitm$:
+The kinding rules also prevent these translational internal term forms  from being well-kinded when $n = 0$. Like the form $\srep{\st}$, these  are retained in values of kind $\kitm$:
 \begin{mathpar}\small
@@
 
-\paragraph{Selectively Abstracted Term Translations} The reason  is again because we will hold these abstract by replacing them with variables.
+\paragraph{Selectively Abstracted Term Translations} The reason for this is again because we will hold argument translations abstract by replacing them with variables.
 The judgement $\edeabs{c}{\argEnv}{\qitm}{\memD}{\memG}{\itm\moutput}{\memD\moutput}{\memG\moutput}$, appearing as the eighth premise of (ana-intro), relates a translational internal term $\qitm$ to an internal term $\itm$ called a \emph{selectively abstracted term translation}, because all references to the translation of an argument (having any type) are replaced with a corresponding variable and all references to the translation of a type constructed by a user-defined tycon other than the ``delegated tycon'' $c$ are replaced with a corresponding abstract type variable as described in Sec.
@@
 Each entry can be read ``argument $n$ having type $\st$ and translation $\iota$ appears as variable $x$ with type $\tau$''.
-For example, \[\small\edeabs{\tcvar{lprod}}{(title; \texttt{"EXMPL 2015"});\Upsilon;\Phi_\text{rstr}\Phi_\text{lprod}}{\qitm_\text{ex}}{\emptyset}{\emptyset}{\iota_\text{ex/abs}}{\memD_\text{ex/abs}}{\memG_\text{ex/abs}}\] where $\itm_\text{ex/abs} := (x_0, (x_1, ()))$, the term translation store is $\memG_\text{ex/abs} := 0 : \stx{title} \leadsto title/x_0 : \alpha_0, 1 : \stx{conf} \leadsto \texttt{"EXMPL 2015"}_\text{IL}/x_1 : \alpha_1$ and the type translation store $\memD_\text{ex/abs}$ is alpha-equivalent to $\memD_\text{paper/abs}$ from Sec.
+For example,
+\[\small\edeabs{\tcvar{lprod}}{(title; \texttt{"EXMPL 2015"});\Upsilon;\Phi_\text{rstr}\Phi_\text{lprod}}{\qitm_\text{ex}}{\emptyset}{\emptyset}{\iota_\text{ex/abs}}{\memD_\text{ex/abs}}{\memG_\text{ex/abs}}\] where $\itm_\text{ex/abs} := (x_0, (x_1, ()))$, the term translation store is $\memG_\text{ex/abs} := 0 : \stx{title} \leadsto title/x_0 : \alpha_0, 1 : \stx{conf} \leadsto \texttt{"EXMPL 2015"}_\text{IL}/x_1 : \alpha_1$ and the type translation store $\memD_\text{ex/abs}$ is alpha-equivalent to $\memD_\text{paper/abs}$ from Sec.
 \ref{sec:selective-type-translation-abstraction}.
@@
 Finally, the twelfth premise checks the abstracted term translation against the abstracted type translation.
-Here, we are checking $\Delta_\text{ex/abs}~\Gamma_\text{ex/abs} \vdash \iota_\text{ex/abs} : \tau_\text{ex/abs}$, i.e.: \[\small(\alpha_0, \alpha_1)~(x_0 : \alpha_0, x_1 : \alpha_1) \vdash (x_0, (x_1, ())) : \alpha_0 \times (\alpha_1 \times \kunit)
-\]
-
+Here, we are checking $\Delta_\text{ex/abs}~\Gamma_\text{ex/abs} \vdash \iota_\text{ex/abs} : \tau_\text{ex/abs}$, i.e.:
+\[\small(\alpha_0, \alpha_1)~(x_0 : \alpha_0, x_1 : \alpha_1) \vdash (x_0, (x_1, ())) : \alpha_0 \times (\alpha_1 \times \kunit)\]
 In other words, the translation of the labeled product $e_\text{ex}$ generated by $\tcvar{lprod}$ is checked with all references to term and type translations of regular strings replaced by variables and type variables, respectively.

### 2014-11-04 att-pldi15/att-pldi15-supplement.tex
Updated TR, related work, cleaned up obdy

@@
 %\documentclass[12pt]{article}
-\documentclass[10pt,preprint]{sigplanconf}
+\documentclass[12pt]{article}
+\usepackage{fullpage,cmu-titlepage2}
+\usepackage{times}
 % The following \documentclass options may be useful:
@@
 \usepackage{mathpartir}
+\usepackage{enumerate}
 \usepackage[usenames,dvipsnames,svgnames,table]{xcolor}
@@
 
-\conferenceinfo{-}{-} 
-\copyrightyear{-} 
-\copyrightdata{[to be supplied]}
-
 %\titlebanner{{\tt \textcolor{Red}{{\small Under Review -- distribute within CMU only.}}}}        % These are ignored unless
@@
 
-\title{Modularly Composing Typed Language Fragments}
-\subtitle{Supplemental Material}
-\authorinfo{}{}{}
+\title{Modularly Composing Typed Language Fragments (Supplemental Material)}
+%\subtitle{Supplemental Material}
+%\authorinfo{}{}{}
 %\authorinfo{Cyrus Omar \and Jonathan Aldrich}
@@
   %         {\{comar, aldrich\}@cs.cmu.edu}
+
+\author{ }
+
+\abstract{
+    This document provides the full technical development described in the paper ``Modularly Composing Typed Language Fragments'' submitted to PLDI 2015. 
+}
+\keywords{Statistical Model Checking, Markov Decision Processes, Reinforcement Learning}
+
+\trnumber{ }
 
@@
 We use {$\mathcal{L}\{{\rightharpoonup}\,{\forall}\,{\mu}\,{\iunit}\,{\times}\,{+}\}$}, the syntax for which is shown in Figure \ref{syntax-IL}, as representative of any intermediate language for a typed functional language.
+
+\begin{figure}[h!]
+\small
+\hspace{-5px}\input{syntax-table-IL}
+\caption{Syntax of {$\mathcal{L}\{{\rightharpoonup}\,{\forall}\,{\mu}\,{\iunit}\,{\times}\,{+}\}$}, our internal language (IL).
+Metavariable $x$ ranges over term variables and $\alpha$ and $t$ both range over type variables.}
+\label{syntax-IL}
+\end{figure}
+
 In fact, our intention is not to prescribe a particular choice of IL, so we will here only review the key metatheoretic properties that the IL must possess.
@@
 If $\emptyset~\emptyset \vdash \iota : \tau$ then either $\iota~\texttt{val}$ or $\iota \mapsto \iota'$.
-\end{theorem}
+\end{lemma}
 \begin{lemma}[Internal Preservation]
@@
 
-\begin{figure}[t]
-\small
-\hspace{-5px}\input{syntax-table-IL}
-\caption{Syntax of {$\mathcal{L}\{{\rightharpoonup}\,{\forall}\,{\mu}\,{\iunit}\,{\times}\,{+}\}$}, our internal language (IL).
-Metavariable $x$ ranges over term variables and $\alpha$ and $t$ both range over type variables.}
-\label{syntax-IL}
-\end{figure}
-
-\begin{figure}
+\begin{figure}[h]
 \small\fbox{$\Delta \vdash \delta : \Delta$}~~$\delta ::=  \emptyset ~|~ \delta, \tau/\alpha$
-
 \begin{mathpar}
@@
 \end{figure}
-\begin{figure}
+
+\begin{lemma}[Internal Type Substitution on Types] If $\Delta \vdash \delta : \Delta'$ and $\Delta\Delta' \vdash \tau$ then $\Delta \vdash [\delta]\tau$.\end{lemma}
+
+\begin{lemma}[Internal Type Substitutions on Typing Contexts] If $\Delta \vdash \delta : \Delta'$ and $\Delta\Delta' \vdash \Gamma$ then $\Delta \vdash [\delta]\Gamma$.
+\end{lemma}
+
+\begin{lemma}[Internal Type Substitutions on Terms] If $\Delta \vdash \delta : \Delta'$ and $\Delta\Delta' \vdash \tau$ and $\Delta\Delta' \vdash \Gamma$ and $\Delta\Delta'~\Gamma \vdash \iota : \tau$ then $\Delta~[\delta]\Gamma \vdash [\delta]\iota : [\delta]\tau$.
+\end{lemma}
+
+\begin{figure}[h]
 \small\fbox{$\Delta~\Gamma \vdash \gamma : \Gamma$}~~$\gamma ::= \emptyset ~|~ \gamma, \iota/x$
@@
 \end{figure}
-\begin{lemma}[Internal Type Substitution on Types] If $\Delta \vdash \delta : \Delta'$ and $\Delta\Delta' \vdash \tau$ then $\Delta \vdash [\delta]\tau$.\end{lemma}
-
-\begin{lemma}[Internal Type Substitutions on Typing Contexts] If $\Delta \vdash \delta : \Delta'$ and $\Delta\Delta' \vdash \Gamma$ then $\Delta \vdash [\delta]\Gamma$.
-\end{lemma}
-
-\begin{lemma}[Internal Type Substitutions on Terms] If $\Delta \vdash \delta : \Delta'$ and $\Delta\Delta' \vdash \tau$ and $\Delta\Delta' \vdash \Gamma$ and $\Delta\Delta'~\Gamma \vdash \iota : \tau$ then $\Delta~[\delta]\Gamma \vdash [\delta]\iota : [\delta]\tau$.
-\end{lemma}
 
@@
 
-\section{Tycon Contexts}
-\begin{figure}\fbox{$\vdash \Phi$}\vspace{-25px}
+\section{Tycons}
+\begin{figure}[h]
+\small
+$\begin{array}{lrcl}
+\textbf{tycons} & c & ::= & \rightharpoonup ~|~ \tc ~|~ \keyw{other}[m]\\
+\textbf{tycon contexts} & \Phi & ::= & \cdot ~|~ \Phi, \tcdef{\tc}{\psi}{\theta}\\
+\textbf{tycon structures} & \theta & ::= & \tcstruct{\st}{\omega} \\
+\textbf{opcon structures} & \omega & ::= & \tcstructn{\st} ~|~ \tcstructc{\omega}{op}{\st}\\
+\textbf{tycon sigs} & \psi & ::= & \tcsig{\kappa}{\chi}\\
+\textbf{opcon sigs}& \chi & ::= & \introsig{\kappa} ~|~ \opsigS{\chi}{op}{\kappa}\\
+\end{array}$
+\caption{Syntax of tycons. Metavariables $\tc$ and $\opname{op}$ range over user-defined tycon and opcon names, respectively, and $m$ ranges over natural numbers.}
+\label{syntax-TC}
+\end{figure}
+
+\subsection{Tycon Contexts}
+Tycon contexts are ordered mappings from tycon names $\tc$ to tycon definitions. The tycon context well-definedness judgement $\vdash \Phi$ is specified in Figure \ref{fig:tycon-ctxs}.
+
+\begin{figure}[h]
+\fbox{$\vdash \Phi$}\vspace{-25px}
 \begin{mathpar}
@@
 \end{figure}
-\begin{figure}[t]
+
+Opcon structures are checked against the tycon's signature.
+
+\begin{figure}[h]
 \small\fbox{$\vdash_\Phi \omega \sim \psi$}
@@
 }{
-    \vdash_\Phi \tcstructc{\omega}{op}{\stx{def}} \sim \tcsig{\ktyidx}{\chi, \opsig{\opname{op}}{\klitidx}}
+    \vdash_\Phi \tcstructc{\omega}{op}{\stx{def}} \sim \tcsig{\ktyidx}{\chi; \opsig{\opname{op}}{\klitidx}}
 }
@@
 
-\subsection{Tycon Context Well-Definedness}
+For the purposes of the SL, only the type signatures are relevant.
+We thus define  judgements $\vdash \psi$, $\vdash \chi$ and $\vdash \Phi~\mathtt{sigsok}$ in Figure \ref{fig:tycon-and-opcon-signature-well-formedness} for checking this only.
+
+The following lemmas characterize these judgements. 
+\begin{figure}[h]
+\fbox{$\vdash \psi$}\vspace{-25px}
+\begin{mathpar}
+\small
+\inferrule[tcsig-ok]{
+    \keq{\emptyset}{\ktyidx}\\
+    \vdash \chi
+}{
+    \vdash \tcsig{\ktyidx}{\chi}
+}
+\end{mathpar}
+\fbox{$\vdash \chi$}\vspace{-25px}
+\begin{mathpar}
+\small
+\inferrule[ocsig-intro-ok]{
+    \emptyset \vdash \kappa
+}{
+    \vdash \introsig{\kappa}
+}
+
+\inferrule[ocsig-targ-ok]{
+    \emptyset \vdash \kappa\\
+    \opname{op} \notin \text{dom}(\chi)
+}{
+    \vdash \opsigS{\chi}{op}{\kappa}
+}
+\end{mathpar}
+\fbox{$\vdash \Phi~\mathtt{sigsok}$}\vspace{-25px}
+\begin{mathpar}
+\small
+\inferrule[tcc-sigsok-emp]{ }{
+    \vdash \emptyset~\mathtt{sigsok}
+}
+
+\inferrule[tcc-sigsok-ext]{
+    \vdash \Phi~\mathtt{sigsok}\\
+    \vdash \psi
+}{
+    \vdash \Phi, \tcdef{\tc}{\psi}{\theta}~\mathtt{sigsok}
+}
+\end{mathpar}
+\caption{Tycon and Opcon Signature Well-Formedness}
+\label{fig:tycon-and-opcon-signature-well-formedness}
+\end{figure}
+
+\begin{lemma}[Inversion on Tycon Context Well-Definedness]
+\label{lemma:inversion-on-tycon-context-well-definedness}
+If (1) $$\vdash \Phi, \tcdef{\tc}{\tcsig{\ktyidx}{\chi}}{\tcstruct{\stx{schema}}{\omega}}$$ then (a) $\vdash \Phi$ and (b) $\tc \notin \text{dom}(\Phi)$ and (c) $\keq{\emptyset}{\ktyidx}$ and (d) $\sofkz{\emptyset}{\emptyset}{\Phi}{\stx{schema}}{\ktyidx \rightarrow \kity}$ and (e) $\vdash_{\Phi,  \tcdef{\tc}{\tcsig{\ktyidx}{\chi}}{\tcstruct{\stx{schema}}{\omega}}} \omega \sim \tcsig{\ktyidx}{\chi}$.
+\end{lemma}
+\begin{proof}
+Rule induction on assumption (1). Rule (tcc-emp) does not apply. Rule (tcc-ext) applies. The conclusions (a-e) are the five premises of (tcc-ext) and thus follow immediately.
+\end{proof}
+
+\begin{lemma}[Tycon Context Composition]
+If (i) $\vdash \Phi_1$ and (ii) $\vdash \Phi_2$ and (iii) $\text{dom}(\Phi_1) \cap \text{dom}(\Phi_2) = \emptyset$ then (a) $\vdash \Phi_1\Phi_2$.
+\end{lemma}
+\begin{proof}
+By syntactic case analysis on $\Phi_1$ and $\Phi_2$.
+
+If $\Phi_1=\emptyset$ and $\Phi_2=\emptyset$, then $\Phi_1\Phi_2=\emptyset$ and (a) follows by (tcc-emp).
+
+If $\Phi_2=\emptyset$ then $\Phi_1\Phi_2=\Phi_1$ and (a) follows by (1).
+
+If $\Phi_1=\emptyset$ then $\Phi_1\Phi_2=\Phi_2$ and (a) follows by (2).
+
+If $\Phi_1=\Phi_1', \tcdef{\tc_1}{\psi_1}{\theta_1}$ and $\Phi_2=\Phi_2',\tcdef{\tc_2}{\psi_2}{\theta_2}$ then $\Phi_1\Phi_2=\Phi_1'\Phi_2',\tcdef{\tc_1}{\psi_1}{\theta_1},\tcdef{\tc_2}{\psi_2}{\theta_2}$.
+Let (1-5) be the result of applying Lemma \ref{lemma:inversion-on-tycon-context-well-definedness} on (i), and (6-10) be the result of applying Lemma \ref{lemma:inversion-on-tycon-context-well-definedness} on (ii).
+By (iii) and the usual properties of domains of finite mappings, we have that (11) $\text{dom}(\Phi_1') \cap \text{dom}(\Phi_2') = \emptyset$ and further by (2) we have that (12) $\tc\notin\keyw{dom}(\Phi_1'\Phi_2')$ and by (8) that (13) $\tc'\notin\text{dom}({\Phi_1'\Phi_2'})$.
+By the IH on (1), (6) and (11) we have that (14) $\vdash \Phi_1'\Phi_2'$.
+By (tcc-ext) on (14), (12), (3), (4) and \todo{weakening of kinding and opcon well-definedness} on (5), we have that (15) $\vdash \Phi_1'\Phi_2', \tcdef{\tc_1}{\psi_1}{\theta_1}$.
+By (tcc-ext) on (15), (13), (8), (9) and SAME on (10), we have (a).
+\end{proof}
+
+\begin{lemma}[Intro Opcon Existence and Well-Definedness]
+If (1) $\vdash_\Phi \omega \sim \tcsig{\ktyidx}{\chi}$ then (a) $\keyw{intro}[\klitidx] \in \chi$ and (b) $\emptyset \vdash \klitidx$ and (c) $\keyw{ana~intro}=\stx{def}\in\omega$ and (d) $\sofkz{\emptyset}{\emptyset}{\Phi}{\stx{def}}{\ktyidx \rightarrow \klitidx \rightarrow \kargs \rightarrow \kitm}$.
+\end{lemma}
+\begin{proof}
+Rule induction on assumption (1).
+If rule (ocstruct-intro) applies then conclusion (c) follows syntactically and (a), (b) and (d) are the three premises and thus follow immediately.
+If rule (ocstruct-targ) applies, then we apply the IH to the first premise.
+\end{proof}
+
+\begin{lemma}[Targeted Opcon Well-Definedness and Unicity]
+If (1) $\vdash_\Phi \omega \sim \tcsig{\ktyidx}{\chi}$ and (2) $\keyw{syn}~\opname{op}=\stx{def} \in \omega$ then (a) $\opname{op}[\klitidx] \in \chi$ and (b) $\emptyset \vdash \klitidx$ and (c) $
+    \sofkn{\emptyset}{\emptyset}{\Phi}{0}{\stx{def}}{\ktyidx \rightarrow \klitidx \rightarrow \kargs \rightarrow (\kty \times \kitm)}$ and (d) if $\keyw{syn}~\opname{op}=\stx{def}' \in \omega$ then $\stx{def}'=\stx{def}$ and (e) if $\opname{op}[\klitidx']\in\chi$ then $\klitidx'=\klitidx$.
+\end{lemma}
+\begin{proof}
+Rule induction on assumption (1).
+Rule (ocstruct-intro) does not apply by assumption (2).
+Rule (ocstruct-targ) applies.
+Conclusion (a) follows syntactically.
+Conclusions (b) and (c) are premises 3 and 4.
+Conclusion (d) follows because either $\stx{def}'=\stx{def}$ (i.e.
+we considering the current definition) or we can apply the IH to premise 1 and the assumption of (d).
+Conclusion (e) follows because premise 2 checks the unicity condition directly.
+\end{proof}
+
+\subsection{Full Examples}
+\subsubsection{Regular Strings}
+
+\subsubsection{Labeled Products}
+
+\subsubsection{Records}
+
+\section{Static Language}
+\subsection{Kind Formation}
+
+\begin{figure}[h]
+\small\fbox{$\kDelta \vdash \kappa$}
+\begin{mathpar}
+\inferrule[kf-arrow]{
+    \kDelta \vdash \kappa_1\\
+    \kDelta \vdash \kappa_2
+}{
+    \kDelta \vdash \karrow{\kappa_1}{\kappa_2}
+}
+
+\inferrule[kf-alpha]{
+    \kalpha \in \kDelta
+}{
+    \kDelta \vdash \kalpha
+}
+
+\inferrule[kf-forall]{
+    \kDelta, \kalpha \vdash \kappa
+}{
+    \kDelta \vdash \kforall{\kalpha}{\kappa}
+}
+
+\inferrule[kf-k]{
+    \kvar \in \kDelta
+}{
+    \kDelta \vdash \kvar
+}
+
+\inferrule[kf-ind]{
+    ???
+}{
+    \kDelta \vdash \kmu{\kvar}{\kappa}
+}
+
+\inferrule[kf-unit]{ }{
+    \kDelta \vdash \kunit
+}
+
+\inferrule[kf-prod]{
+    \kDelta \vdash \kappa_1\\
+    \kDelta \vdash \kappa_2
+}{
+    \kDelta \vdash \kprod{\kappa_1}{\kappa_2}
+}
+
+\inferrule[kf-sum]{\kDelta \vdash \kappa_1\\
+    \kDelta \vdash \kappa_2
+}{
+    \kDelta \vdash \ksum{\kappa_1}{\kappa_2}
+}
+
+\inferrule[kf-ty]{ }{\kDelta \vdash \kty}
+
+\inferrule[kf-ity]{ }{\kDelta \vdash \kity}
+
+\inferrule[kf-itm]{ }{\kDelta \vdash \kitm}
+\end{mathpar}
+\caption{Kind Formation}
+\label{fig:kind-formation}
+\end{figure}
+\begin{lemma}[Kind Variable Substitution - Kinds]~\\
+\begin{enumerate}
+\item If $\kDelta, \kalpha \vdash \kappa$ and $\kDelta \vdash \kappa'$ then $\kDelta \vdash [\kappa'/\kalpha]\kappa$.
+\item If $\kDelta, \kvar \vdash \kappa$ and $\kDelta \vdash \kappa'$ then $\kDelta \vdash [\kappa'/\kvar]\kappa$.
+\end{enumerate}
+\end{lemma}
+\begin{figure}[h]
+???
+\caption{Positive Kinds}
+\label{fig:positive-kinds}
+\end{figure}
+
+\begin{figure}[h]
+\small\fbox{$\kDelta \vdash \kGamma$}
+\begin{mathpar}
+\inferrule[kctx-emp]{ }{\kDelta \vdash \emptyset}
+
+\inferrule[kctx-ext]{
+    \kDelta \vdash \kGamma\\
+    \kDelta \vdash \kappa
+}{
+    \kDelta \vdash \kGamma, \sx :: \kappa
+}
+\end{mathpar}
+\caption{Kinding Context Formation}
+\label{fig:kinding-context-formation}
+\end{figure}
 
@@
 
-\subsection{Full Examples}
-
-\section{Static Language}
-\subsection{Kind Formation}
-
-\subsection{Kinding Context Formation}
+\begin{lemma}[Equality Kind Well-Formedness]
+If $\keq{\emptyset}{\kappa}$ then $\emptyset \vdash \kappa$. 
+\end{lemma}
 
@@
 
-\subsection{Dynamic Semantics}
+\subsection{Static Dynamics}
 
-\subsection{Kind Safety}
+\begin{lemma}[Static Canonical Forms]
+...
+\end{lemma}
 
-\section{Types}
+Discussion of decidability and weak normalization.
+
 \subsection{Type Translations}
@@
 
+\subsection{Arguments}
+\begin{definition}[Argument Interface Kind]
+The kind abbreviated $\karg$ is defined as\[\karg := (\kty \rightarrow \kitm) \times (\kunit \rightarrow \kty\times\kitm)\]
+\end{definition}
+
+\subsection{Kind Safety}
+\paragraph{Kind Safety} Kind safety ensures that normalization of well-kinded static terms cannot go wrong. We can take a standard progress and preservation based approach. 
+\begin{theorem}[Static Progress]\label{thm:static-progress}
+If $\sofkn{\emptyset}{\emptyset}{\Phi}{n}{\st}{\kappa}$ and $\vdash \Phi$ and $|\es|=n$ and $\vdash_\Phi \Upsilon \leadsto \Gamma$ then $\sval{\st}{\es; \Upsilon; \Phi}$ or $\serr{\st}{\es; \Upsilon; \Phi}$ or $\sstep{\st}{\es; \Upsilon; \Phi}{\st'}$.
+\end{theorem}
+
+\begin{theorem}[Static Preservation]\label{thm:static-preservation}
+If $\sofkn{\emptyset}{\emptyset}{\Phi}{n}{\st}{\kappa}$ and $\vdash \Phi$ and $|\es|=n$ and $\vdash_\Phi \Upsilon \leadsto \Gamma$ and $\sstep{\st}{\es; \Upsilon; \Phi}{\st'}$ then $\sofkn{\emptyset}{\emptyset}{\Phi}{n}{\st'}{\kappa}$.
+\end{theorem}
+The case in the proof of Theorem \ref{thm:static-preservation} for $\ssyn{n}$ requires that the following theorem be mutually defined.
+The mutual induction is well-founded because the total number of argument lists in the terms being considered decreases.
+\begin{theorem}[Type Synthesis]
+If $\vdash \Phi$ and $\vdash_\Phi \Upsilon \leadsto \Gamma$ and $\esynX{e}{\st}{\iota}$ then $\vdash_\Phi \st \leadsto \tau$ (and thus $\istype{\st}{\Phi}$). 
+\end{theorem}
+
+\section{External Language}
+\subsection{Additional Desugarings}
+\subsection{Typing}
 \paragraph{Unicity}
@@
 
-\section{External Language}
-\subsection{Additional Desugarings}
-\subsection{Typing}
 \subsection{Proof of Regular String Soundness Tycon Invariant}
@@
 We assume that the definitions of the standard helper functions $\small\svar{nil} :: \kforall{\kalpha}{\klist{\kalpha}}$ and $\small\svar{rcons} :: \kforall{\kalpha}{\klist{\kalpha} \rightarrow \kalpha \rightarrow \klist{\kalpha}}$, which adds an item to the end of a list, have been substituted into these rules.
-The result is that the $n$th element of the argument interface list simply wraps the static terms $\sana{n}{\st}$ and $\ssyn{n}$.\end{document}
+The result is that the $n$th element of the argument interface list simply wraps the static terms $\sana{n}{\st}$ and $\ssyn{n}$.
+
+\begin{lemma}
+\small If $\sofkn{\kDelta}{\kGamma}{\Phi}{n'}{\st}{\kappa}$ and $n > n'
+$ then $\sofkn{\kDelta}{\kGamma}{\Phi}{n}{\st}{\kappa}$.
+\end{lemma}
+
+\end{document}

### 2014-11-04 att-pldi15/att-pldi15.tex
Updated TR, related work, cleaned up obdy

@@
 \textbf{tycon sigs} & \psi & ::= & \tcsig{\kappa}{\chi}\\
-\textbf{opcon sigs}& \chi & ::= & \introsig{\kappa} ~|~ \opsigS{\theta}{op}{\kappa}\\
-\end{array}\vspace{-3px}$
+\textbf{opcon sigs}& \chi & ::= & \introsig{\kappa} ~|~ \opsigS{\chi}{op}{\kappa}\\
+\end{array}$
 \caption{Syntax of tycons. Metavariables $\tc$ and $\opname{op}$ range over user-defined tycon and opcon names, respectively, and $m$ ranges over natural numbers.}
@@
 %The supplement gives the analagous rules.
+
 The final two forms of translational internal term are $\anatrans{n}{\st}$ and $\syntrans{n}$.
@@
 \lambda\svar{tyidx}{:}\klist{\klbl\times\kty}.\lambda\svar{tmidx}{:}\klist{\klbl}.\lambda\svar{args}{:}\kargs.\\
-\quad \keyw{let}~\svar{inhabited}:\kunit=\svar{uniqmap}~\svar{tyidx}\\
+\quad \keyw{let}~\svar{inhabited}:\kunit=\svar{uniqmap}~\svar{tyidx}~\keyw{in}\\
 \quad \skap{\kitm}{\skap{\karg}{\skap{\klbl}{\skap{\klbl\times\kty}{\svar{listrec3}}}}}~\svar{tyidx}~\svar{tmidx}~\svar{args}~\sqitm{\itriv}\\
 \quad\quad \lambda \svar{rowtyidx}{:}\klbl\times\kty.\lambda \svar{rowtmidx}{:}\klbl.\lambda\svar{rowarg}{:}\karg.\lambda \svar{r}{:}\kitm.\\
-\quad\quad\quad \keyw{letpair}~(\svar{rowlbl}, \svar{rowty})=\svar{rowtyidx}\\
-\quad\quad\quad \keyw{let}~\svar{lok}::\kunit=\svar{lbleq}~\svar{rowlbl} ~\svar{rowtmidx}\\
-\quad\quad\quad \keyw{let}~\svar{rowtr} :: \kitm = \svar{ana}~\svar{rowarg}~\svar{rowty}\\
+\quad\quad\quad \keyw{letpair}~(\svar{rowlbl}, \svar{rowty})=\svar{rowtyidx} ~\keyw{in}\\
+\quad\quad\quad \keyw{let}~\svar{lok}::\kunit=\svar{lbleq}~\svar{rowlbl} ~\svar{rowtmidx}~\keyw{in}\\
+\quad\quad\quad \keyw{let}~\svar{rowtr} :: \kitm = \svar{ana}~\svar{rowarg}~\svar{rowty}~\keyw{in}\\
 \quad\quad\quad \sqitm{(\quq{\svar{rowtr}}, \quq{\svar{r}})
@@
 
-The \emph{argument interface list} is then simply a static value of kind $\kargs$ of length $n$ where the $i$th entry is the argument interface $(\slam{\kty}{\svar{ty}}{\sana{i}{\svar{ty}}}, \slam{\kunit}{\_}{\ssyn{i}})$.
+The \emph{argument interface list} is a static list of length $n$ where the $i$th entry is $(\slam{\kty}{\svar{ty}}{\sana{i}{\svar{ty}}}, \slam{\kunit}{\_}{\ssyn{i}})$.
 It is generated by the judgement $\keyw{args}(n)=\stx{args}$, where $n$ is the length of the argument list, written $|\es|=n$.
@@
 \end{mathpar}
-The following lemma characterizes thus characterizes the argument list interface generation judgement:
+The following lemma characterizes thus characterizes the argument list interface generation judgement:\vspace{-3px}
 \begin{lemma}
 \small If $\keyw{args}(n)= \stx{args}$ then $\sofkn{\emptyset}{\emptyset}{\Phi}{n}{\stx{args}}{\kargs}$.
-\end{lemma}
-
+\end{lemma}\vspace{-3px}
 The rule (ocstruct-intro) ruled out writing either of these forms explicitly in an opcon definition by checking against the bound $n=0$.
 This is to prevent out-of-bounds errors: tycon providers do not write these forms directly, only accessing them via the argument interface list, which is guaranteed to have the correct length.
-This is  safe because $n$ is only an upper bound:
-\begin{lemma}
-\small If $\sofkn{\kDelta}{\kGamma}{\Phi}{n'}{\st}{\kappa}$ and $n > n'
-$ then $\sofkn{\kDelta}{\kGamma}{\Phi}{n}{\st}{\kappa}$.
-\end{lemma}
+This is  safe because $n$ is only an upper bound, so it can be relaxed safely.
 
@@
 
-The selectively abstracted term translation judgement proceeds recursively along shared forms, like the selectively abstracted type translation judgement.
-The key rule for references to argument translations derived via analysis is below ($\syntrans{n}$ is analagous; see supplement):
+This judgement proceeds recursively along shared forms, like the selectively abstracted type translation judgement.
+The key rule for references to argument translations derived via analysis is below ($\syntrans{n}$ is analagous):
 \begin{mathpar}\small
@@
 \[\small(\alpha_0, \alpha_1)~(x_0 : \alpha_0, x_1 : \alpha_1) \vdash (x_0, (x_1, ())) : \alpha_0 \times (\alpha_1 \times \kunit)\]
-In other words, the translation of the labeled product $e_\text{ex}$ generated by $\tcvar{lprod}$ is checked with all references to term and type translations of regular strings replaced by variables and type variables, respectively.
-Because our intro opcon definition treated arguments parametrically, the check succeeds.
-We  describe  an ill-behaved operator in Sec.
+In other words, the translation of the labeled product $e_\text{ex}$ generated by $\tcvar{lprod}$ is checked with the references to term and type translations of regular strings replaced by variables and type variables, respectively.
+But because our  definition treated arguments parametrically, the check succeeds.% We  describe  an ill-behaved operator in Sec.
 \ref{metatheory}.
@@
 
-The remaining premises of (syn-targ) are analagous to the corresponding premises in (ana-intro), with the only difference being that we check the selectively abstracted.
-term translation against the selectively abstracted type translation of the synthesized type, but the delegated tycon is that of the type synthesized by the target argument.
+The remaining premises of (syn-targ) are analagous to the corresponding premises in (ana-intro), with the only difference being that we check the selectively abstracted  term translation against the selectively abstracted type translation of the synthesized type, but the delegated tycon is that of the type synthesized by the target argument.
 
@@
 The case in the proof of Theorem \ref{thm:static-preservation} for $\ssyn{n}$ requires that the following theorem be mutually defined.
-The mutual induction is well-founded because the total number of argument lists in the terms being considered decreases.
+%The mutual induction is well-founded because the total number of intro and targeted sub-terms being considered decreases.
 \begin{theorem}[Type Synthesis]
@@
 Type safety in a typed translation semantics requires that well-typed external terms translate to well-typed internal terms.
-Type safety for the IL \cite{pfpl} then implies that well-typed external terms cannot go wrong.
-To prove this, we must actually prove a stronger theorem, \emph{type-preserving translation}, analagous to type-preserving compilation in the typed compilation literature \cite{tarditi+:til-OLD}:% produced as the translation of the external type.
+Type safety for the IL \cite{pfpl} then implies that evaluation cannot go wrong.
+To prove this, we must prove a stronger theorem, \emph{type-preserving translation}, analagous to type-preserving compilation in the typed compilation literature \cite{tarditi+:til-OLD}:% produced as the translation of the external type.
 Note that we need only to state our top-level theorems for the analysis judgement  because of the subsumption rule.
 
-\begin{theorem}[Type Preserving Translation]
+\begin{theorem}[Type-Preserving Translation]
 If $\vdash \Phi$ and $\vdash_\Phi \Upsilon \leadsto \Gamma$ and $\vdash_\Phi \st \leadsto \tau$ and $\eanaX{e}{\st}{\iota}$ then $\emptyset~\Gamma \vdash \iota : \tau$.
@@
 \begin{proof-sketch}
-We induct on the typing judgement.
 The interesting cases are (ana-intro), (ana-intro-other), (syn-trans) and (syn-trans-other); the latter two arise via subsumption.
 The result follows directly from the final premise of each rule, combined with lemmas that state that if $\memD \leadsto \delta : \Delta_\text{abs}$ and $\memG \leadsto \gamma : \Gamma_\text{abs}$ then $\emptyset \vdash \delta : \Delta_\text{abs}$ and $\Delta_\text{abs}~\Gamma \vdash \gamma : \Gamma_\text{abs}$, i.e.
-that all variables in $\Delta_\text{abs}$ and $\Gamma_\text{abs}$ have well-formed/well-typed substitutions in $\delta$ and $\gamma$,  and so applying them in the conclusions of the rules gives a well-typed term.
+that all variables in $\Delta_\text{abs}$ and $\Gamma_\text{abs}$ have well-formed/well-typed substitutions in $\delta$ and $\gamma$,  and so applying them gives a well-typed term.
 \end{proof-sketch}
@@
 \end{tyconinvariant}
-\begin{proof-sketch} The proof is not unusually difficult because we have fixed the tycon context $\Phi_\text{rstr}$, so we can simply treat the calculus like a type-directed compiler for a calculus with only two tycons, $\rightharpoonup$ and $\tcvar{rstr}$.
+\begin{proof-sketch} The proof is not unusually difficult because we have fixed the tycon context $\Phi_\text{rstr}$, so we can essentially treat the calculus like a type-directed compiler for a calculus with only two tycons, $\rightharpoonup$ and $\tcvar{rstr}$.
 Such a calculus and compiler specification was given in \cite{sanitation-psp14}, so we must simply show that the opcon definitions in $\tcvar{rstr}$ adequately capture these specifications using standard techniques for the SL, a simply-typed functional language \cite{conf/pldi/Chlipala07}.
@@
 
-The reason why the rule (syn-targ-other) is never a problem in proving a tycon invariant, i.e.
+The reason why (syn-targ-other) is never a problem in proving a tycon invariant, i.e.
 \emph{translation independence} of tycons, turns out to be the same reason extending the tycon context conserves all tycon invariants.
@@
 $\tc$, the mapping still has a type constructed by that tycon with the same type translation.
-Thus we can apply the original tycon invariant.
-The only condition on $P(\sttyidx, \iota)$ is that it be preserved under this mapping on $\sttyidx$.
+If $P(\sttyidx, \iota)$ is preserved under the mapping over $\sttyidx$ then we can simply invoke the assumed tycon invariant.
 We call such propositions \emph{modular}.
-Non-modular propositions are  uninteresting because the mapping's only effect on static terms is to replace all types constructed by $\tc'$ with a unique type constructed by $\keyw{other}[m]$, for $m$ uniquely associated with $\tc'$.
-On external terms, the mapping replaces all intro and targeted sub-terms associated with $\tc'$ with an equivalent one that passes through the rules (ana-intro-other) and (syn-targ-other) by partially applying the intro and targeted opcon definitions to generate the term indices.
-Typing, kinding, static normalization and the selective translation abstraction judgements are all preserved under this mapping, defined inductively in the supplement.
+Non-modular propositions are uninteresting because the mapping's effect on static terms is only to replace all types constructed by $\tc'$ with a unique type constructed by $\keyw{other}[m]$, for $m$ uniquely associated with $\tc'$.
+Any propositions decidable by the SL are modular.
+On external terms, the mapping replaces all intro and targeted terms associated with $\tc'$ with an equivalent one that passes through the rules (ana-intro-other) and (syn-targ-other) by partially applying the intro and targeted opcon definitions to generate the term indices.
+Typing, kinding, static normalization and selective translation abstraction are preserved under the mapping, defined inductively in the supplement.
 \end{proof-sketch}
@@
 
-\section{Related and Future Work}\label{prior-work}
-As discussed throughout the paper, our treatment of concrete syntax defers to recent work on \emph{type-specific languages} \cite{TSLs}.
-%As in our work, TSLs determine the meaning of introductory terms based on the type the term is analyzed against, but TSLs only support syntactic elaboration to a fixed semantics.
-
+\section{Related Work and Discussion}\label{prior-work}
 %\paragraph{Term Rewriting}
-Language-integrated static term rewriting  (``macro'') systems, like Template Haskell \cite{SheardPeytonJones:Haskell-02} and Scala's static macros \cite{ScalaMacros2013}, can be used to decrease complexity when an isomorphic embedding into the underlying type system is possible, but cannot extend the type system directly.% If each metaprogram is invoked like a function, inner macros are expanded before outer macros, and the mechanism enforces abstraction barriers by ensuring that the rewriting logic is hygienic, cannot modify surrounding code, and does not depend on shared state, then compositional reasoning is possible: the meaning of a term depends only on its subterms, not on the specific position it appears in a program.% Complex macros can, however, still be difficult to reason about, so many static macro systems impose further constraints (e.g.
+Language-integrated static term rewriting systems, like Template Haskell \cite{SheardPeytonJones:Haskell-02} and Scala's static macros \cite{ScalaMacros2013}, can be used to decrease complexity when an isomorphic embedding into the underlying type system is possible.
+% If each metaprogram is invoked like a function, inner macros are expanded before outer macros, and the mechanism enforces abstraction barriers by ensuring that the rewriting logic is hygienic, cannot modify surrounding code, and does not depend on shared state, then compositional reasoning is possible: the meaning of a term depends only on its subterms, not on the specific position it appears in a program.% Complex macros can, however, still be difficult to reason about, so many static macro systems impose further constraints (e.g.
 Scala recommends using only its ``black box'' macros, which  enforce a function-like typing discipline).
@@
 %Rewriting systems that permit global pattern-based dispatch, however, do not admit strong modular reasoning principles (the same term might match multiple patterns, defined separately, creating ambiguities).
+Similarly, when an embedding that preserves a desired static semantics exists, but a different  embedding preserves the cost semantics, term rewriting  can also be used to perform ``optimizations'', achieving an isomorphism.
+Care is needed when this changes the type of a term.
+Type abstraction has been used for  this purpose in \emph{lightweight modular staging} (LMS) \cite{Rompf:2012:LMS}.
+In both cases, the type system is fixed (e.g.
+in LMS, Scala's).
 
@@
 %\paragraph{Optimization}
-When a simple embedding that preserves the static semantics exists, but a different  embedding is needed to preserve a desired cost semantics, term rewriting  can also be used to perform ``optimizations'', achieving an isomorphic embedding.
-Care is needed when this changes the type of a term to ensure that invariants are preserved.
-Type abstraction has been used for this purpose in work on \emph{lightweight modular staging} (LMS) \cite{Rompf:2012:LMS}.
-This does not enable new type systems, i.e.
-in \cite{Rompf:2012:LMS}, Scala's type system must suffice.
+ %This does not enable new type systems, i.e.
+in \cite{Rompf:2012:LMS}, Scala's type system is used.
 %As long as transformations are meaning-preserving (which can also be shown mechanically), they can be applied in any order.
@@
 %When T, the solution is less clear. 
-When new static distinctions are needed within an existing type, but new operators are not necessary and cost is acceptable, one solution is to develop a system of \emph{refinement types} for the language \cite{Freeman91}.
+When new distinctions are needed within an existing type, but new operators are not necessary, one solution is to develop an overlying system of \emph{refinement types} \cite{Freeman91}.
 For example, a refinement of  integers might distinguish negative integers.
-Proposals for ``pluggable type systems'' describe composing such type refinement systems \cite{Brac04a,Andreae:2006:FIP:1167473.1167479}.
-Refinements of abstract types can be used to hold representations abstract.
+Proposals for ``pluggable type systems'' describe composing such   systems \cite{Brac04a,Andreae:2006:FIP:1167473.1167479}.
+Refinements of abstract types can be used for representation independence, but note that the type being refined is not held abstract.
 %Because the dynamic semantics are fixed,  composition is safe.
 Although our regular string types could perhaps be approximated with refinements, this does not permit the advanced operations, optimizations and representation independence results we enable.
-
-%\paragraph{Language Frameworks}
-Our interest is in a more general situation, of which refinement types can be seen in ways as a degenerate case: when introducing a fragment requires defining new types and new operators, where the types and translations arising from the new operators are a  function of  statically valued information.
-For example, labeled products require new opcons that use labels and types, which exist only statically.
-%Regular string types could also support an operation that the internal string type would not (e.g.
+Such a system could be seen in some ways as a degenerate mode of use of our work: we further cover the cases when new  operators are needed. 
+For example, the $\tcvar{lprod}$ opcons use labels and types, which exist only statically; they are not refinements of binary products because no row projection operator exists.
+%Labeled products are not refinements of binary products because the label projection operator is not defined for binary products.%Regular string types could also support an operation that the internal string type would not (e.g.
 extraction of a subgroup defined in the type index).% The regular string type's group projection operator had a specialized cost semantics, so a simple refinement would not be able to approximate it.
@@
 
-Many \emph{language frameworks} and \emph{compiler generators} exist to simplify  dialect implementation (e.g.
-\todo{citations}).
+Many \emph{language frameworks} exist that can simplify  dialect implementation (cf.
+\cite{erdweg2013state}).
 %Some \emph{extensible compilers} are also actually language frameworks, because they permit the introduction of new constructs, not just new meaning-preserving optimizations.
@@
 T
-These either do not support forming languages from separately defined fragments or provide few clear modular reasoning principles that make it possible to reason separately about them.
-\todo{read and describe amal ahmed's recent paper}
-
-Proof assistants and logical frameworks can be used to mechanize dialect specification and metatheory, but they similarly require a complete specification.
-Chlipala has suggested using  proof automation that covers a variety of language features to decrease the burden when forking a  dialect, but this is a fundamentally heuristic approach \cite{Chlipala10}.
-Our approach provides a conservativity theorem.
-No proof terms are needed to guarantee modular reasoning:   mistakes in a tycon definition can only weaken invariants at that tycon.
-%This facilitates exploratory language design and suggests educational applications.
-A promising avenue for future work is to introduce optional proof mechanization into the SL (by making it dependently kinded), and to mechanize the metatheory of @$\lambda$ itself.%Every combination of fragments is a new dialect, and must be reasoned about monolithically.
+These sometimes do not support forming  languages from fragments due to the ``expression problem'' (EP) \cite{wadler1998expression,Reynolds75}.
+We sidestep the most serious consequences of the EP by leaving our abstract syntax entirely fixed, instead  delegating to tycons.
+Few tools require knowledge of all external tycons in a typed translation semantics.
+As discussed previously, our treatment of concrete syntax in both the EL and SL defers to recent work on \emph{type-specific languages}, which takes a similar bidirectional approach for syntactic elaborations \cite{TSLs}.
+Some language frameworks do address the EP, but  provide few clear principles from which to reason modularly about metatheoretic properties specific to typed languages, like type safety and tycon invariants, as discussed in Sec.
+\ref{intro}.
+One important decision that makes this tractable in @$\lambda$ is to associate opcons with tycons, forming \emph{modular tycons}, rather than considering terms and types separately.
+
+Proof assistants can be used to specify and mechanize the metatheory of languages, but also usually require a complete specification (this  has been identified as a key challenge \cite{aydemir05tphols}).
+Techniques for composing specifications and proofs exist \cite{conf/popl/DelawareOS13,Delaware11,conf/plpv/SchwaabS13}, but they  require additional proofs at composition-time and provide no guarantees that \emph{all} fragments having some modularly checkable property can safely and  conservatively be composed, as in our work.
+%For example, Delaware et al.
+develop a technique based on software product lines that requires specifying assumptions at feature boundaries and providing proofs of these for each composition \cite{Delaware11}.
+In later work based on Mendler-style $f$-algebras, this problem of \emph{feature interactions} persists for many lemmas, including those related to tycon invariants (e.g.
+canonical forms) \cite{conf/popl/DelawareOS13}.
+Both there and in work by Chlipala  \cite{Chlipala10}, the suggestion is to use  proof automation, but this is a heuristic approach.
+
+Fragment providers need not develop mechanized proofs with our approach.
+Under a fixed tycon context, the calculus can be reasoned about like a very small type-directed compiler \cite{tarditi+:til-OLD,conf/pldi/Chlipala07}.
+Errors in an opcon definition lead to failure when typechecking external terms that use them.
+Incorrect opcon definitions (relative to a specification, e.g.
+\cite{fulton-thesis} for regular strings) can only weaken expected invariants at that tycon, like incorrectly implemented modules in ML, so tycons can reasonably be tested in real code without worrying about non-local impact.
+The simplicity of opcons also suggests educational applications (like TinkerType \cite{LevinPierce99}).
+To eliminate even these localized failure modes for ``reliability-critical'' tycons, we are working on \emph{optional} proof mechanization in the SL (by making it dependently kinded), and on mechanizing the metatheory of @$\lambda$ itself.%Every combination of fragments is a new dialect, and must be reasoned about monolithically.
 %Efforts to reason modularly areeed, several problems can come up when fragments are combined, so .
@@
 
-Type abstraction is, encouragingly, also the technique underlying modular reasoning in ML-like languages.
+Type abstraction, encouragingly, also underlies modular reasoning in ML-like languages \cite{pfpl,harper1997programming,conf/oopsla/AminRO14} and languages with other forms of ADTs \cite{liskov1974programming}.
 Indeed, proofs of tycon invariants can rely on the same parametricity theorem \cite{WadlerThms}.
-Our work has technical similarities to work on  translating modules to System $\mathbf{F}_\omega$ \cite{conf/tldi/RossbergRD10}.
-Unlike in module systems, type translations (analagous to the choice of representation for an abstract type) can be \emph{computed} based on a type index.
-Moreover, there can be arbitrarily many operators because they arise by providing a term index to an opcon.
-Their semantics can be complex because a static function computes the type and translation.
-In contrast, modules can only specify a fixed number of operations, and each must have function type.
-Note that although we did not specify quantification over external types here for simplicity, we conjecture that it is complementary (and thus @$\lambda$ could serve as the core of a language with a ML-style module system).
-Another direction is \emph{tycon functors}, which abstract over tycons with the same  signature.
+Our work is technically reminiscent of  work on  translating modules to System $\mathbf{F}_\omega$ \cite{conf/tldi/RossbergRD10}.
+Unlike in module systems, type translations (analagous to the choice of representation for an abstract type) are \emph{statically computed} based on a type index, rather than declared directly.
+Moreover, there can be arbitrarily many operators because they arise by providing a term index to an opcon, and their  semantics can be complex because a static function computes types and translations.
+In contrast, modules and ADTs can only specify a fixed number of operations, and each must have function type.
+Note also that these are not competing mechanisms: we did not specify quantification over external types here for simplicity, but we conjecture that it is complementary and thus @$\lambda$ could serve as the core of a language with an ML-style module system.
+Another related direction is \emph{tycon functors}, which would abstract over tycons with the same signature to support  different cost models.
 % This permits us tocode targeting a separate IL, and permit operations whose types cannot simply be written as functions.
@@
 
-One limitation of our approach is that it supports only  fragments with the same ``shape'' of typing judgement.
+One limitation of our approach is that it supports only  fragments with the standard ``shape'' of typing judgement.
 Fragments that require new forms of contexts (e.g.
-symbol contexts \cite{pfpl})  cannot be defined.
-The language controls variable binding, so new notions of binding, e.g.
-linear type systems, cannot be defined.
-Another limitation is that opcons cannot invoke one another (e.g.
-a ``length'' opcon on regular strings could not construct a natural number).
-We conjecture that these are not fundamental limitations.
+symbol contexts \cite{pfpl}), for example, cannot be defined.
+Relatedly, the language controls variable binding, so, for example, linear type systems, cannot be defined.
+Another limitation is that opcons cannot directly invoke one another (e.g.
+a \opname{len} opcon on regular strings could not construct a natural number).
+We conjecture that these are not fundamental limitations and expect @$\lambda$ to serve as a minimal foundation for future efforts that increase  expressiveness while maintaining the fundamental guarantees established  here.
 %(tycons could define new contexts which are threaded opaquely through ``outside'' operations).

### 2014-11-04 att-pldi15/att-pldi15[Conflict].tex
Updated PLDI by removing some details of SL and stuff

--- 2014-11-04 att-pldi15/att-pldi15.tex
@@
 universal quantification over types).
-We use {$\mathcal{L}\{{\rightharpoonup}\,{\forall}\,{\mu}\,{\iunit}\,{\times}\,{+}\}$}, the syntax for which is shown in Figure \ref{syntax-IL}, as representative of any intermediate language for a typed functional language.
+We use {$\mathcal{L}\{{\rightharpoonup}\,{\forall}\,{\mu}\,{\iunit}\,{\times}\,{+}\}$}, the syntax for which is shown in Figure \ref{syntax-IL}, as representative of a typical intermediate language for a typed functional language.
 %Note that the IL has a fixed semantics; changes to the IL induce different dialects of @$\lambda$.
@@
 The multi-step judgement $\iota \mapsto^{*} \iota\moutput$ is the reflexive, transitive closure of the stepping judgement and the evaluation judgement $\iota \Downarrow \iota'$ is defined iff $\iota \mapsto^{*} \iota'$ and $\iota'~\mathtt{val}$.
-Both the static and dynamic semantics of the IL can be found in any standard textbook covering typed lambda calculi (we directly follow \cite{pfpl}), so we assume familiarity.
+Both the static and dynamic semantics of the IL can be found in any standard textbook covering typed lambda calculi (we directly follow \cite{pfpl}), so we assume familiarity and omit the details.
 
@@
 The syntax of the SL is given in Figure \ref{syntax-SL}.
-The portion of the SL covered by the first row of kinds and first four rows of static terms in the syntax forms an entirely standard  functional programming language consisting of total functions, universal quantification over kinds, inductive kinds (constrained by the standard positivity condition to prevent non-termination), and products and sums.
+The portion of the SL covered by the first row of kinds and first four rows of static terms in the syntax forms an entirely standard  functional programming language consisting of total functions, universal quantification over kinds, inductive kinds, and products and sums.
 The reader can consider these as forming a total subset of ML or a simply-typed subset of Coq.
@@
 
-\paragraph{Examples}
-Our first example of a user-defined tycon is $\tcvar{rstr}$.
-It has index kind $\krx$, which classifies static regular expression patterns (defined as an inductive sum kind in the usual way).
-Types constructed by $\tcvar{rstr}$ will classify \emph{regular strings}, which statically obey the invariant that they are in the regular language specified by the type index \cite{sanitation-psp14}.
+\paragraph{Examples} Two examples of tycon contexts (each defining only one tycon)  and their signatures are shown in Figure \ref{fig:example-tycons}.
+
+Our first example, $\tcvar{rstr}$, has index kind $\krx$, which classifies static regular expression patterns (defined as an inductive sum kind in the usual way).
+Types constructed by $\tcvar{rstr}$ will classify \emph{regular strings}, which obey the invariant that they are in the regular language specified by the type index \cite{sanitation-psp14}.
 For example, $\stx{title} := \sty{\tcvar{rstr}}{\concrx{.+}}$ will classify non-empty strings and $\stx{conf} := \sty{\tcvar{rstr}}{\concrx{[A-Z]+ \digit\digit\digit\digit}}$ will classify conference names.
@@
 In the tycon context containing both tycon definitions, $\Phi_\text{rstr}\Phi_\text{lprod}$, we can define a labeled product type classifying conference papers, $\stx{paper} := \sty{\tcvar{lprod}}{\{\conclbl{title} : \stx{title}, \conclbl{conf} : \stx{conf}\}}$.
-Note that $\istype{\stx{paper}}{\Phi_\text{rstr}\Phi_\text{lprod}}$ and we again use kind-specific syntax, in this case for $\klbl$ and $\klist{\klbl \times \kty}$.
+Note that $\istype{\stx{paper}}{\Phi_\text{rstr}\Phi_\text{lprod}}$ and we again use kind-specific syntax, in this case for $\klist{\klbl \times \kty}$.
 
 \subsection{Static Type Case Analysis}
-\begin{figure}\hfill \fbox{$\vdash \Phi$}\vspace{-25px}\begin{mathpar}
-\small
-\inferrule[tcc-ext]{
-    \vdash \Phi\\
-    \tc \notin \text{dom}(\Phi)\\
-    \keq{\emptyset}{\ktyidx}\\
-    \sofkz{\emptyset}{\emptyset}{\Phi}{\stx{schema}}{\ktyidx \rightarrow \kity}\\\\
-    \vdash_{\Phi,  \tcdef{\tc}{\tcsig{\ktyidx}{\chi}}{\tcstruct{\stx{schema}}{\omega}}} \omega \sim \tcsig{\ktyidx}{\chi}
-}{
-    \vdash \Phi, \tcdef{\tc}{\tcsig{\ktyidx}{\chi}}{\tcstruct{\stx{schema}}{\omega}}
-}\vspace{-10px}
-\end{mathpar}
-\caption{Tycon context well-definedness. We omit the trivial case for when $\Phi=\cdot$ for concision.}
-\label{fig:tycon-ctxs}
-\vspace{-8px}
+
+\begin{figure}\small
+\hspace{-10px}\begin{tabular}{ll}
+$\begin{array}{l}\Phi_\text{rstr} := \tcdef{\tcvar{rstr}}{\psi_\text{rstr}}{\\
+\quad\tcstruct{\stx{rstr/trans}}{\\
+\quad\tcstructn{\stx{rstr/intro}}};\\
+\quad\keyw{syn}~\opname{conc} = \stx{rstr/conc};\\
+\quad\keyw{syn}~\opname{case} = \stx{rstr/case};\\
+\quad\cdots\\
+}\end{array}$ & $\begin{array}{l}\Phi_\text{lprod} := \tcdef{\tcvar{lprod}}{\psi_\text{lprod}}{\\
+\quad\tcstruct{\stx{lprod/trans}}{\\
+\quad\tcstructn{\stx{lprod/intro}}};\\
+\quad\keyw{syn}~\opname{\#} = \stx{lprod/prj};\\
+\quad\keyw{syn}~\opname{conc} = \stx{lprod/conc};\\
+\quad\cdots\\
+}\end{array}$
+\end{tabular}\vspace{3px}\\
+$\psi_\text{rstr} := \tcsig{\krx}{\introsig{\keyw{Str}}; \opname{conc}[\kunit]; \opname{case}[\keyw{StrPattern}]; ...}$\\
+$\psi_\text{lprod} := \tcsig{\klist{\klbl\times\kty}}{\introsig{\klist{\klbl}}; \opname{\#}[\klbl]; \opname{conc}[\kunit]; ...}$
+\caption{Example tycon contexts and signatures.}
+\label{fig:example-tycons}\vspace{-8px}
 \end{figure}
@@
 The dynamics (see supplement) are straightforwardly consistent with the intuition above.% We will see an example of its use in Sec. \ref{sec:targops}.
+
 \subsection{Tycon Context Well-Definedness}
+
+\begin{figure}\hfill \fbox{$\vdash \Phi$}\vspace{-25px}\begin{mathpar}
+\small
+\inferrule[tcc-ext]{
+    \vdash \Phi\\
+    \tc \notin \text{dom}(\Phi)\\
+    \keq{\emptyset}{\ktyidx}\\
+    \sofkz{\emptyset}{\emptyset}{\Phi}{\stx{schema}}{\ktyidx \rightarrow \kity}\\\\
+    \vdash_{\Phi,  \tcdef{\tc}{\tcsig{\ktyidx}{\chi}}{\tcstruct{\stx{schema}}{\omega}}} \omega \sim \tcsig{\ktyidx}{\chi}
+}{
+    \vdash \Phi, \tcdef{\tc}{\tcsig{\ktyidx}{\chi}}{\tcstruct{\stx{schema}}{\omega}}
+}\vspace{-10px}
+\end{mathpar}
+\caption{Tycon context well-definedness. We omit the trivial case for when $\Phi=\cdot$ for concision.}
+\label{fig:tycon-ctxs}
+\vspace{-8px}
+\end{figure}
 The tycon context well-definedness judgement, $\vdash \Phi$, shown in Figure \ref{fig:tycon-ctxs}, requires that all tycon names are unique and performs additional checks,  described below.
@@
 \end{mathpar}
-\caption{Checking opcon structures against tycon signatures.}
+\caption{Opcon structure kinding against tycon signatures}
 \label{ocstruct}\vspace{-5px}
@@
 As discussed previously, our treatment of concrete syntax in both the EL and SL defers to recent work on \emph{type-specific languages}, which takes a similar bidirectional approach for syntactic elaborations \cite{TSLs}.
-Some language frameworks do address the EP, but  provide few clear principles from which to reason modularly about metatheoretic properties specific to typed languages, like type safety and tycon invariants, as discussed in Sec.
-\ref{intro}.
-One important decision that makes this tractable in @$\lambda$ is to associate opcons with tycons, forming \emph{modular tycons}, rather than considering terms and types separately.
+Some language frameworks do address the EP, but  provide few clear principles from which to reason modularly about metatheoretic properties specific to typed languages, like type safety and tycon invariants.
+One important decision that makes this tractable in @$\lambda$ is to associate opcons with tycons, forming \emph{modular tycons}, rather than considering terms and types both as open datatypes \cite{conf/ppdp/LohH06}.
 
@@
 Under a fixed tycon context, the calculus can be reasoned about like a very small type-directed compiler \cite{tarditi+:til-OLD,conf/pldi/Chlipala07}.
-Errors in an opcon definition lead to failure when typechecking external terms that use them.
+Errors in an opcon definition lead to failure to typecheck, a form of \emph{translation validation} \cite{Pnueli-Siegel-Singerman98}.
 Incorrect opcon definitions (relative to a specification, e.g.
-\cite{fulton-thesis} for regular strings) can only weaken expected invariants at that tycon, like incorrectly implemented modules in ML, so tycons can reasonably be tested in real code without worrying about non-local impact.
+\cite{sanitation-psp14} for regular strings) can at worst weaken expected invariants at that tycon, like incorrectly implemented modules in ML, so tycons can reasonably be tested in real code without worrying about non-local impact.
 The simplicity of opcons also suggests educational applications (like TinkerType \cite{LevinPierce99}).
@@
 Our work is technically reminiscent of  work on  translating modules to System $\mathbf{F}_\omega$ \cite{conf/tldi/RossbergRD10}.
-Unlike in module systems, type translations (analagous to the choice of representation for an abstract type) are \emph{statically computed} based on a type index, rather than declared directly.
+Unlike in module systems, type translations (analagous to the choice of representation for an abstract type) are statically \emph{computed} based on a type index, rather than statically \emph{declared}.
 Moreover, there can be arbitrarily many operators because they arise by providing a term index to an opcon, and their  semantics can be complex because a static function computes types and translations.
@@
 Note also that these are not competing mechanisms: we did not specify quantification over external types here for simplicity, but we conjecture that it is complementary and thus @$\lambda$ could serve as the core of a language with an ML-style module system.
-Another related direction is \emph{tycon functors}, which would abstract over tycons with the same signature to support  different cost models.
+Another related direction is \emph{tycon functors}, which would abstract over tycons with the same signature to support  different cost semantics.
 % This permits us tocode targeting a separate IL, and permit operations whose types cannot simply be written as functions.
@@
 Another limitation is that opcons cannot directly invoke one another (e.g.
-a \opname{len} opcon on regular strings could not construct a natural number).
-We conjecture that these are not fundamental limitations and expect @$\lambda$ to serve as a minimal foundation for future efforts that increase  expressiveness while maintaining the fundamental guarantees established  here.
+a \opname{len} opcon on regular strings could not construct a natural number; see supplement for a workaround, however).
+We conjecture that these are not fundamental limitations and expect @$\lambda$ to serve as a minimal foundation for future efforts that increase  expressiveness while maintaining fundamental guarantees like type safety and conservativity as established  here.
 %(tycons could define new contexts which are threaded opaquely through ``outside'' operations).

### 2014-11-04 att-pldi15/att-pldi15.tex
Updated

@@
 \AtBeginDocument{%
- \abovedisplayskip=5pt
+ \abovedisplayskip=4pt
  \abovedisplayshortskip=4pt
@@
 }
+
+\usepackage[compact]{titlesec}  
+\titlespacing{\section}{0pt}{6pt}{2pt}
+\titlespacing{\subsection}{0pt}{3pt}{0pt}
+\titlespacing{\subsubsection}{0pt}{3pt}{0pt}
+\titlespacing{\paragraph}{0pt}{3pt}{5pt}
 
@@
 In this paper, we begin from first principles by specifying a language, @$\lambda$, in the style of many full-scale languages: as a bidirectionally typed translation semantics.
-Only the $\rightarrow$ type constructor (tycon) is built in; all other external tycons (we show a variant on record types and constrained string types) are defined by extending a \emph{tycon context}.
+Only the $\rightarrow$ type constructor (tycon) is built in; all other external tycons (we show constrained strings and a variant on records) are defined by extending a \emph{tycon context}.
 Each tycon defines the semantics of its  term-level operators (e.g.
@@
 $\mathcal{L}\{{\rightharpoonup}\,{\forall}\,{\mu}\,{\keyw{1}}\,{\times}\,{+}\}$ is the language of partial function types, polymorphic types, recursive types, nullary and binary product types and binary sum types (its syntax is shown in Figure \ref{syntax-IL}, discussed below).
-In the literature, researchers often describe fragments using a simple calculus having a ``catch-all'' constant and base type to stand notionally for all other terms and types that may also be included in some future complete language.
+Another common practice is to describe a fragment using a simple calculus having a ``catch-all'' constant and base type to stand notionally for all other terms and types that may also be included in some future complete language (e.g.
+\cite{sanitation-psp14}).
 
@@
 %Indeed, the proliferation of language dialects constructed for this reason might be taken as an evidence that the core language is not as ``general'', when considered comprehensively, as might be hoped.
-Each time a fragment must be exposed directly, rather than as a library, a new \emph{dialect} of a language must be constructed.
+Each time a fragment must be exposed directly, rather than as a library, a new \emph{dialect} of the language must be constructed.
 Within the ML lineage, for example, dialects that go beyond ``core ML'' abound:
@@
 %\verb|sprintf| in the OCaml dialect statically distinguishes format strings from strings.% All of these are implemented as dialects of existing languages, presumably because a strong encoding was not feasible.
-
+\vspace{-3px}
 \item
@@
 a safe FFI, item 3 above, is needed between every pair of dialects).
+And as just mentioned, composing dialects is non-trivial.
 
@@
 For example, a module defining sets in ML can hold the representation of sets abstract, ensuring that any invariants maintained by the functions in the module (e.g.
-uniqueness, if using a list representation) will hold no matter which other modules are separately in use by a client \cite{harper1997programming}.
-Other langauges offer similar facilities, often grouped under the moniker of \emph{abstract data types} \cite{liskov1974programming}.
+uniqueness, if using a list representation) will hold no matter which other modules are in use by a client \cite{harper1997programming}.
+Other languages provide similar facilities, e.g.
+Scala's abstract type members \cite{conf/oopsla/AminRO14}.
 %Mechanisms that can help decrease the complexity of an embedding without violating abstraction barriers are thus valuable, and we will lead into our work in Sec.
@@
 
-\paragraph{Contributions} In this paper, we take foundational steps towards this goal by  constructing a simple but surprisingly powerful core calculus, @$\lambda$ (the ``actively typed'' lambda calculus).
+\paragraph{Contributions} In this paper, we take foundational steps toward this goal by  constructing a simple but surprisingly powerful core calculus, @$\lambda$ (the ``actively typed'' lambda calculus).
 %Despite its minimality, it can host a variety of practical semantic extensions like those described above, while maintaining strong metatheoretic guarantees and, crucially, providing modular reasoning properties.
@@
 Our main focus here is on semantic extensions.
-The static and dynamic semantics are given simultaneously as a \emph{bidirectionally typed translation semantics}, i.e.
+The static and dynamic semantics are specified simultaneously as a \emph{bidirectionally typed translation semantics}, i.e.
 the key judgements take the form:\[\esynX{e}{\st\moutput}{\iota\moutput} ~~~~~\text{and}~~~~~ \eanaX{e}{\st}{\iota\moutput}\]
@@
 Note that our specifications in this paper are intended to be algorithmic: we indicate ``outputs'' when introducing judgement forms by \emph{mode annotations}, $\moutput$; these are not part of the judgement's syntax.
-In particular, note that the type is an ``output'' only in the synthetic judgement.
+In particular, note that the type is an ``output'' only for the synthetic judgement.
 
@@
 
-External typing contexts $\Upsilon$ map variables to types, so we also need the judgement $\vdash_\Phi \Upsilon \leadsto \Gamma$, where $\Gamma$ maps variables to the corresponding type translation.
+External typing contexts $\Upsilon$ map variables to types, so we also need the judgement $\vdash_\Phi \Upsilon \leadsto \Gamma$ to construct a $\Gamma$ that maps the same variables to corresponding type translations.
 
@@
 %Internal terms, $\iota$, together with internal types, $\tau$, form the \emph{typed internal language}. 
-@$\lambda$ relies on a typed internal language supporting type abstraction (i.e.
-universal quantification over types).
-We use {$\mathcal{L}\{{\rightharpoonup}\,{\forall}\,{\mu}\,{\iunit}\,{\times}\,{+}\}$}, the syntax for which is shown in Figure \ref{syntax-IL}, as representative of any intermediate language for a typed functional language.
+@$\lambda$ requires a typed internal language supporting type abstraction (i.e.
+universal quantification over types) \cite{Reynolds94anintroduction}.
+We use {$\mathcal{L}\{{\rightharpoonup}\,{\forall}\,{\mu}\,{\iunit}\,{\times}\,{+}\}$}, the syntax for which is shown in Figure \ref{syntax-IL}, as representative of a typical intermediate language for a typed functional language.
 %Note that the IL has a fixed semantics; changes to the IL induce different dialects of @$\lambda$.
@@
 The syntax of the SL is given in Figure \ref{syntax-SL}.
-The portion of the SL covered by the first row of kinds and first four rows of static terms in the syntax forms an entirely standard  functional programming language consisting of total functions, universal quantification over kinds, inductive kinds (constrained by the standard positivity condition to prevent non-termination), and products and sums.
-The reader can consider these as forming a total subset of ML or a simply-typed subset of Coq.
-The semantics also directly follow \cite{pfpl}, so we  omit the details of this core here.
-Only three new  kinds are needed for the SL to serve its role as the language used to   control typechecking and translation of the EL: $\kty$ (Sec.
+The portion of the SL covered by the first row of kinds and first row of static terms, some of which are elided for concision, forms an entirely standard  functional programming language consisting of total functions, universal quantification over kinds, inductive kinds, and products and sums.
+The reader can consider these as forming a total subset of ML or a simply-typed subset of Coq and we will assume standard conveniences from such languages (e.g.
+let bindings, pattern matching and inference of type parameters when applying polymorphic functions) are available in examples for concision.
+The semantics we assume also directly follows \cite{pfpl}, so we also omit the details of this core here.
+Only three new  kinds are needed for the SL to serve its role as the language used to   control typing and translation of the EL: $\kty$ (Sec.
 \ref{types}), $\kity$ (Sec.
@@
 \caption{Kinding rules for types, which take the form $\sty{c}{\sttyidx}$ where $c$ is a tycon and $\sttyidx$ is the type index.}
-\label{fig:types}\vspace{-8px}
+\label{fig:types}\vspace{-10px}
 \end{figure*}
@@
 
-\paragraph{Examples}
-Our first example of a user-defined tycon is $\tcvar{rstr}$.
-It has index kind $\krx$, which classifies static regular expression patterns (defined as an inductive sum kind in the usual way).
-Types constructed by $\tcvar{rstr}$ will classify \emph{regular strings}, which statically obey the invariant that they are in the regular language specified by the type index \cite{sanitation-psp14}.
+\paragraph{Examples} Two examples of tycon contexts (each defining only one tycon)  and their signatures are shown in Figure \ref{fig:example-tycons}.
+\footnote{In examples, we omit leading $\emptyset$, used as the base case for finite mappings, and $\cdot$, used as the base case for finite sequences, for concision.}
+
+Our first example, $\tcvar{rstr}$, has index kind $\krx$, which we assume classifies static regular expression patterns (defined as an inductive sum kind in the usual way).
+Types constructed by $\tcvar{rstr}$ will classify terms that behave as \emph{regular strings}, statically obeying the invariant that they are in the regular language specified by the type index \cite{sanitation-psp14}.
 For example, $\stx{title} := \sty{\tcvar{rstr}}{\concrx{.+}}$ will classify non-empty strings and $\stx{conf} := \sty{\tcvar{rstr}}{\concrx{[A-Z]+ \digit\digit\digit\digit}}$ will classify conference names.
 The type indices are  here written using standard concrete syntax for concision; recent work has specified how to define type-specific (or here, kind-specific) syntax like this composably in libraries \cite{TSLs}.
-We define the tycon context containing only the definition of $\tcvar{rstr}$, $\Phi_\text{rstr} := \tcdef{\tcvar{rstr}}{\tcsig{\krx}{\chi_\text{rstr}}}{\theta_\text{rstr}}$.\footnote{In examples, we omit leading $\emptyset$, used as the base case for finite mappings, and $\cdot$, used as the base case for finite sequences, for concision.
-}
+%We define the tycon context containing only the definition of $\tcvar{rstr}$, $\Phi_\text{rstr} := \tcdef{\tcvar{rstr}}{\tcsig{\krx}{\chi_\text{rstr}}}{\theta_\text{rstr}}$.
 
@@
 We choose the index kind of $\tcvar{lprod}$ to be $\klist{\klbl \times \kty}$, where list kinds are defined as inductive sums in the usual way, and $\klbl$ classifies static representations of row labels.
-The tycon context containing only $\tcvar{lprod}$'s definition is $\small\Phi_\text{lprod} := \tcdef{\tcvar{lprod}}{\tcsig{\klist{\klbl \times \kty}}{\chi_\text{lprod}}}{\theta_\text{lprod}}$.
 In the tycon context containing both tycon definitions, $\Phi_\text{rstr}\Phi_\text{lprod}$, we can define a labeled product type classifying conference papers, $\stx{paper} := \sty{\tcvar{lprod}}{\{\conclbl{title} : \stx{title}, \conclbl{conf} : \stx{conf}\}}$.
-Note that $\istype{\stx{paper}}{\Phi_\text{rstr}\Phi_\text{lprod}}$ and we again use kind-specific syntax, in this case for $\klbl$ and $\klist{\klbl \times \kty}$.
+Note that $\istype{\stx{paper}}{\Phi_\text{rstr}\Phi_\text{lprod}}$ and we again use kind-specific syntax, in this case for $\klist{\klbl \times \kty}$.
 
 \subsection{Static Type Case Analysis}
-\begin{figure}\hfill \fbox{$\vdash \Phi$}\vspace{-25px}\begin{mathpar}
-\small
-\inferrule[tcc-ext]{
-    \vdash \Phi\\
-    \tc \notin \text{dom}(\Phi)\\
-    \keq{\emptyset}{\ktyidx}\\
-    \sofkz{\emptyset}{\emptyset}{\Phi}{\stx{schema}}{\ktyidx \rightarrow \kity}\\\\
-    \vdash_{\Phi,  \tcdef{\tc}{\tcsig{\ktyidx}{\chi}}{\tcstruct{\stx{schema}}{\omega}}} \omega \sim \tcsig{\ktyidx}{\chi}
-}{
-    \vdash \Phi, \tcdef{\tc}{\tcsig{\ktyidx}{\chi}}{\tcstruct{\stx{schema}}{\omega}}
-}\vspace{-10px}
-\end{mathpar}
-\caption{Tycon context well-definedness. We omit the trivial case for when $\Phi=\cdot$ for concision.}
-\label{fig:tycon-ctxs}
-\vspace{-8px}
+
+\begin{figure}\small
+\hspace{-10px}\begin{tabular}{ll}
+$\begin{array}{l}\Phi_\text{rstr} := \tcdef{\tcvar{rstr}}{\psi_\text{rstr}}{\\
+\quad\tcstruct{\stx{rstr/trans}}{\\
+\quad\tcstructn{\stx{rstr/intro}}};\\
+\quad\keyw{syn}~\opname{conc} = \stx{rstr/conc};\\
+\quad\keyw{syn}~\opname{case} = \stx{rstr/case};\\
+\quad\cdots}\end{array}$ & $\begin{array}{l}\Phi_\text{lprod} := \tcdef{\tcvar{lprod}}{\psi_\text{lprod}}{\\
+\quad\tcstruct{\stx{lprod/trans}}{\\
+\quad\tcstructn{\stx{lprod/intro}}};\\
+\quad\keyw{syn}~\opname{\#} = \stx{lprod/prj};\\
+\quad\keyw{syn}~\opname{conc} = \stx{lprod/conc};\\
+\quad\cdots}\end{array}$
+\end{tabular}\vspace{3px}\\
+$\psi_\text{rstr} := \tcsig{\krx}{\introsig{\keyw{Str}}; \opname{conc}[\kunit]; \opname{case}[\keyw{StrPattern}]; ...}$\\
+$\psi_\text{lprod} := \tcsig{\klist{\klbl\times\kty}}{\introsig{\klist{\klbl}}; \opname{\#}[\klbl]; \opname{conc}[\kunit]; ...}$
+\caption{Example tycon contexts and signatures.}
+\label{fig:example-tycons}\vspace{-8px}
 \end{figure}
@@
 The dynamics (see supplement) are straightforwardly consistent with the intuition above.% We will see an example of its use in Sec. \ref{sec:targops}.
+
 \subsection{Tycon Context Well-Definedness}
-The tycon context well-definedness judgement, $\vdash \Phi$, shown in Figure \ref{fig:tycon-ctxs}, requires that all tycon names are unique and performs additional checks,  described below.
-
+
+% \begin{figure}\hfill \fbox{$\vdash \Phi$}\vspace{-25px}
+% \caption{Tycon context well-definedness. We omit the trivial case for when $\Phi=\cdot$ for concision.}
+% \label{fig:tycon-ctxs}
+% \vspace{-8px}
+% \end{figure}
+The tycon context well-definedness judgement, $\vdash \Phi$, requires that all tycon names are unique and performs three additional checks,  described below (we omit (tcc-emp)).
+\begin{mathpar}
+\small
+\inferrule[tcc-ext]{
+    \vdash \Phi\\
+    \tc \notin \text{dom}(\Phi)\\
+    \keq{\emptyset}{\ktyidx}\\
+    \sofkz{\emptyset}{\emptyset}{\Phi}{\stx{schema}}{\ktyidx \rightarrow \kity}\\\\
+    \vdash_{\Phi,  \tcdef{\tc}{\tcsig{\ktyidx}{\chi}}{\tcstruct{\stx{schema}}{\omega}}} \omega \sim \tcsig{\ktyidx}{\chi}
+}{
+    \vdash \Phi, \tcdef{\tc}{\tcsig{\ktyidx}{\chi}}{\tcstruct{\stx{schema}}{\omega}}
+}
+\end{mathpar}
 \subsection{Type Equivalence}\label{sec:type-equivalence}
@@
 For example, our schema for $\tcvar{rstr}$ will simply choose to ignore the type index and translate regular strings to strings, of internal type abbreviated $\keyw{str}$.
-We abbreviate the corresponding translational internal type $\hat{\keyw{str}}$ and define $\small\theta_\text{rstr} := \tcstruct{\slam{\krx}{\svar{tyidx}}{\sqity{\hat{\keyw{str}}}}}{\omega_\text{rstr}}$.
-
-The kinding and dynamics for these shared forms simply proceed recursively, e.g.
+We abbreviate the corresponding translational internal type $\hat{\keyw{str}}$ and define $\stx{rstr/trans} := \slam{\krx}{\svar{tyidx}}{\sqity{\hat{\keyw{str}}}}$.
+The kinding and dynamics for shared forms proceed recursively, e.g.
 \begin{mathpar}
@@
 \end{mathpar}
-These forms are needed in the translation schema for $\tcvar{lprod}$, which generates nested binary product types by recursing over the type index and referring to the translations of the types therein (this is not the only viable choice, e.g.
-we could also have used a list).
-We assume $\small\svar{listrec} :: \kforall{\kalpha_1}{\kforall{\kalpha_2}{\klist{\kalpha_1}\rightarrow\kalpha_2\rightarrow(\kalpha_1\rightarrow\kalpha_2\rightarrow\kalpha_2)\rightarrow\kalpha_2}}$ in defining $\theta_\text{lprod} := \keyw{trans}=\stx{lprod/trans}~\keyw{in}~\omega_\text{lprod}$  where $\stx{lprod/trans}:=$
+These forms are needed in the translation schema for $\tcvar{lprod}$, which generates nested binary product types (though we could also have used a list) by recursing over the type index and referring to the translations of the types therein.
+We assume the standard $\small\svar{listrec} :: \kforall{\kalpha_1}{\kforall{\kalpha_2}{\klist{\kalpha_1}\rightarrow\kalpha_2\rightarrow(\kalpha_1\rightarrow\kalpha_2\rightarrow\kalpha_2)\rightarrow\kalpha_2}}$ in defining $\stx{lprod/trans}:=$
 \[\small
@@
 \end{array}\]
-Applying this translation schema to the index of $\stx{paper}$, for example, produces the value $\stx{paper/trans} := \sqity{\qity_\text{paper/trans}}$ where $\small\qity_\text{paper/trans} := \srep{\stx{title}}\times(\srep{\stx{conf}}\times \iunit)$.
-Note that we do not include logic to optimize away the trailing unit type for simplicity (and to again emphasize that many translations are possible for any given type).%For simplicity, and to make the point that the choice of representation has only tycon-local implications, our translation schema does not optimize away
+Applying this translation schema to the index of $\stx{paper}$, for example, produces  $\stx{paper/trans} := \sqity{\qity_\text{paper/trans}}$ where $\small\qity_\text{paper/trans} := \srep{\stx{title}}\times(\srep{\stx{conf}}\times \iunit)$.
+Note that our schema did not remove the trailing unit type for simplicity (and to again emphasize that this choice will have only local implications due to the \emph{translation independence} property we will enforce as the core of our modularity guarantee).%For simplicity, and to make the point that the choice of representation has only tycon-local implications, our translation schema does not optimize away
 
 \subsubsection{Selective Type Translation Abstraction}\label{sec:selective-type-translation-abstraction}
-References to type translations are maintained in values of kind $\kity$ like this to  allow us to selectively hold them abstract, which will be the key to our main results.
+References to type translations are maintained in values of kind $\kity$ like this to  allow us to selectively hold them abstract, which will be the key to translation independence.
 This can be thought of as analagous to the process in ML by which the true identity of an abstract type in a module is held abstract outside the module until after typechecking.
 The judgement $\small\tdeabs{\Phi}{c}{\qity}{\memD}{\ity\moutput}{\memD\moutput}$ relates a normalized translational internal type $\qity$ to an internal type $\tau$, called a \emph{selectively abstracted type translation} because references to translations of types constructed by a tycon other than the ``delegated tycon'', $c$, are replaced by a corresponding type variable, $\alpha$.
-The \emph{type translation store} $\memD ::= \emptyset ~|~ \memD, \st \leftrightarrow \ity/\alpha$ maintains the correspondence between types, their actual translations and the unique type variables which appear in their place, e.g.
+The \emph{type translation store} $\memD ::= \emptyset ~|~ \memD, \st \leftrightarrow \ity/\alpha$ maintains this correspondence between types, their actual translations and the unique type variables which appear in their place, e.g.
 $\small\tdeabs{\Phi_\text{rstr}\Phi_\text{lprod}}{\tcvar{lprod}}{\qity_\text{paper/trans}}{\emptyset}{\tau_\text{paper/abs}}{\memD_\text{paper/abs}}$
@@
 The rules for the selective type translation abstraction judgement are straightforward.
-We first recurse generically over sub-terms of $\qity$ until sub-terms of form $\srep{\st}$ are encountered (see supplement).
+We recurse generically over sub-terms of $\qity$ until sub-terms of form $\srep{\st}$ are encountered (see supplement).
 The translation of partial function types is direct and is not held abstract, so that lambdas can be used as the sole binding construct in the EL:
@@
 \end{mathpar}
-The translation of a user-defined type constructed by the delegated tycon is determined by calling the translation schema and checking that the type translation it generates refers only to type variables generated from $\memD'$:
+The translation of a user-defined type constructed by the delegated tycon is determined by calling the translation schema and checking that the type translation it generates refers only to type variables in $\memD'$:
 \begin{mathpar}
@@
 The function has access to the type index, the term index and an interface to the list of arguments, and is kind is checked by the  judgement $\vdash_\Phi \omega \sim \psi$, which appeared as the final premise of the rule (tcc-ext) and is defined in Figure \ref{ocstruct}.
-
-For example, the opcon structure of $\tcvar{rstr}$ is  $\small\omega_\text{rstr} := \tcstructn{\stx{rstr/intro}}; \omega_\text{rstr/targops}$ where $\small\stx{rstr/intro}:=$
+For example, $\small\stx{rstr/intro}:=$
 \[\small\begin{array}{l}
@@
 %The supplement gives the analagous rules.
-
 The final two forms of translational internal term are $\anatrans{n}{\st}$ and $\syntrans{n}$.
 These stand in for the translation of argument $n$, the first if it arises via analysis against type $\st$ and the second if it arises via type synthesis.
-Before giving the rules, let us motivate the mechanism with the intro opcon definition for $\tcvar{lprod}$.
-We have $\omega_\text{lprod} := \tcstructn{\stx{lprod/intro}}; \omega_\text{lprod/targops}$ where $\stx{lprod/intro} :=$
+Before giving the rules, let us motivate the mechanism with the intro opcon definition for $\tcvar{lprod}$, shown in Figure \ref{fig:lprod-intro}.
+
+\begin{figure}\vspace{-10px}
 \[\small
@@
 \quad\quad\quad \sqitm{(\quq{\svar{rowtr}}, \quq{\svar{r}})
-}\end{array}
-\]
+}\end{array}\]\vspace{-10px}
+\caption{The intro opcon definition for $\tcvar{lprod}$.}
+\label{fig:lprod-intro}\vspace{-8px}
+\end{figure}
 
@@
 We chose not to formalize these for simplicity, and to demonstrate this general technique.
-%An analagous technique could be used to implement record types by requiring that the index be sorted (see supplement for additional discussion).%, leaving indices where the labels did not appear sorted uninhabited.
+An analagous technique could be used to implement record types by requiring that the index be sorted (see supplement).%, leaving indices where the labels did not appear sorted uninhabited.
 In both cases, we could provide a static helper function of kind $\klist{\klbl\times\kty}\rightarrow \kty$ that checked well-formedness immediately before constructing the requested type.
@@
 Then, we request type analysis of the corresponding argument, $\svar{rowarg}$,  against the type in the type index, $\svar{rowty}$, by writing $\svar{ana}~\svar{rowarg}~\svar{rowty}$.
-Here, $\svar{ana}$ is a helper function defined below that produces a translational internal term  of the form $\sqitm{\anatrans{n}{\st}}$, where $n$ is the position of $\svar{rowarg}$ in $\svar{args}$ and $\st$ is the value of $\svar{rowty}$, if the argument is well-typed, and raises an error if not.
+Here, $\svar{ana}$ is a helper function defined below that triggers type analysis of the provided argument, producing a translational internal term  of the form $\sqitm{\anatrans{n}{\st}}$, where $n$ is the position of $\svar{rowarg}$ in $\svar{args}$ and $\st$ is the value of $\svar{rowty}$, upon success or  raising an error if not.
 The final line constructs a nested tuple based on this translation and the recursive result.
@@
 
-\paragraph{Argument Interface Lists} We define $\karg$, the kind of \emph{argument interfaces}, as a simple product of functions, $\karg := (\kty \rightarrow \kitm) \times (\kunit \rightarrow \kty\times\kitm)$, one for analysis and the other for synthesis, both having the expected kind for these operations.
-The helper functions $\svar{ana}$ and $\svar{syn}$ simply project the corresponding function out, $\svar{ana} := \slam{\karg}{\svar{arg}}{\keyw{fst}(\svar{arg})}$ and $\svar{syn}  :=  \slam{\karg}{\svar{arg}}{\keyw{snd}(\svar{arg})}$.
+\paragraph{Argument Interface Lists} We define the kind of \emph{argument interfaces} as a simple product of functions, $\karg := (\kty \rightarrow \kitm) \times (\kunit \rightarrow \kty\times\kitm)$, one for analysis and the other for synthesis The helper functions $\svar{ana}$ and $\svar{syn}$ simply project the corresponding function out, $\svar{ana} := \slam{\karg}{\svar{arg}}{\keyw{fst}(\svar{arg})}$ and $\svar{syn}  :=  \slam{\karg}{\svar{arg}}{\keyw{snd}(\svar{arg})}$.
 
@@
 
- Recall that the kinding judgement is indexed by $n$, which is an upper bound on the argument index of terms of the form $\sana{n}{\st}$ and $\ssyn{n}$. This is enforced in their kinding rules:
+ Recall that the kinding judgement is indexed by $n$. This is an upper bound on the argument index of terms of the form $\sana{n}{\st}$ and $\ssyn{n}$. This is enforced in their kinding rules:
 \begin{mathpar}\small
@@
 \end{mathpar}
-The following lemma characterizes thus characterizes the argument list interface generation judgement:\vspace{-3px}
-\begin{lemma}
-\small If $\keyw{args}(n)= \stx{args}$ then $\sofkn{\emptyset}{\emptyset}{\Phi}{n}{\stx{args}}{\kargs}$.
-\end{lemma}\vspace{-3px}
+Thus, if $\keyw{args}(n)= \stx{args}$ then $\sofkn{\emptyset}{\emptyset}{\Phi}{n}{\stx{args}}{\kargs}$.
+
 The rule (ocstruct-intro) ruled out writing either of these forms explicitly in an opcon definition by checking against the bound $n=0$.
 This is to prevent out-of-bounds errors: tycon providers do not write these forms directly, only accessing them via the argument interface list, which is guaranteed to have the correct length.
-This is  safe because $n$ is only an upper bound, so it can be relaxed safely.
+%This is  safe because $n$ is only an upper bound, so it can be relaxed safely.
 
@@
 \paragraph{Selectively Abstracted Term Translations} The reason for this is again because we will hold argument translations abstract by replacing them with variables.
-The judgement $\edeabs{c}{\argEnv}{\qitm}{\memD}{\memG}{\itm\moutput}{\memD\moutput}{\memG\moutput}$, appearing as the eighth premise of (ana-intro), relates a translational internal term $\qitm$ to an internal term $\itm$ called a \emph{selectively abstracted term translation}, because all references to the translation of an argument (having any type) are replaced with a corresponding variable and all references to the translation of a type constructed by a user-defined tycon other than the ``delegated tycon'' $c$ are replaced with a corresponding abstract type variable as described in Sec.
-\ref{sec:selective-type-translation-abstraction}.
+The judgement $\edeabs{c}{\argEnv}{\qitm}{\memD}{\memG}{\itm\moutput}{\memD\moutput}{\memG\moutput}$, appearing as the eighth premise of (ana-intro), relates a translational internal term $\qitm$ to an internal term $\itm$ called a \emph{selectively abstracted term translation}, because all references to the translation of an argument (having any type) are replaced with a corresponding variable and, as in Sec.
+\ref{sec:selective-type-translation-abstraction}, all references to the translation of a type constructed by a user-defined tycon other than the ``delegated tycon'' $c$ are replaced with a corresponding abstract type variable.
 The type translation store, $\memD$, discussed previously, and term translation store, $\memG$, track these correspondences.
 Term translation stores have  syntax $\memG ::= \emptyset ~|~ \memG, n : \st \leadsto \iota/x : \tau$.
-Each entry can be read ``argument $n$ having type $\st$ and translation $\iota$ appears as variable $x$ with type $\tau$''.
-For example,
-\[\small\edeabs{\tcvar{lprod}}{(title; \texttt{"EXMPL 2015"});\Upsilon;\Phi_\text{rstr}\Phi_\text{lprod}}{\qitm_\text{ex}}{\emptyset}{\emptyset}{\iota_\text{ex/abs}}{\memD_\text{ex/abs}}{\memG_\text{ex/abs}}\] where $\itm_\text{ex/abs} := (x_0, (x_1, ()))$, the term translation store is $\memG_\text{ex/abs} := 0 : \stx{title} \leadsto title/x_0 : \alpha_0, 1 : \stx{conf} \leadsto \texttt{"EXMPL 2015"}_\text{IL}/x_1 : \alpha_1$ and the type translation store $\memD_\text{ex/abs}$ is alpha-equivalent to $\memD_\text{paper/abs}$ from Sec.
+Each entry can be read ``argument $n$ having type $\st$ and translation $\iota$ appears as variable $x$ with type $\tau$'', e.g.
+\[\small\edeabs{\tcvar{lprod}}{(title; \texttt{"EXMPL 2015"});\Upsilon;\Phi_\text{rstr}\Phi_\text{lprod}}{\qitm_\text{ex}}{\emptyset}{\emptyset}{\iota_\text{ex/abs}}{\memD_\text{ex/abs}}{\memG_\text{ex/abs}}\] where $\itm_\text{ex/abs} := (x_0, (x_1, ()))$ and $\memG_\text{ex/abs} := 0 : \stx{title} \leadsto title/x_0 : \alpha_0, 1 : \stx{conf} \leadsto \texttt{"EXMPL 2015"}_\text{IL}/x_1 : \alpha_1$ and  $\memD_\text{ex/abs}$ is $\memD_\text{paper/abs}$ from Sec.
 \ref{sec:selective-type-translation-abstraction}.
 
-This judgement proceeds recursively along shared forms, like the selectively abstracted type translation judgement.
-The key rule for references to argument translations derived via analysis is below ($\syntrans{n}$ is analagous):
+This judgement proceeds recursively along shared forms, like the selectively abstracted type translation judgement in Sec.
+\ref{sec:selective-type-translation-abstraction}.
+The key rule for references to argument translations derived via analysis is below ($\syntrans{n}$ is analagous; the full rules are in the supplement).
+Note that we are rederiving the translation already determined in (s-ana-success) for simplicity (in practice, this might be cached):
 \begin{mathpar}\small
@@
 In this case $\small\tau_\text{ex/abs} := \alpha_0 \times (\alpha_1 \times \kunit)$ (alpha-equivalent to $\tau_\text{paper/abs}$ in Sec.
-\ref{sec:selective-type-translation-abstraction}), and the eleventh premise extracts a type substitution, $\delta_\text{ex/abs}$, and type formation context, $\Delta_\text{ex/abs}$, from $\memD_\text{ex/abs}$, again equivalent to $\delta_\text{conf/abs}$ and $\Delta_\text{conf/abs}$ from Sec.
+\ref{sec:selective-type-translation-abstraction}).
+The eleventh premise extracts a type substitution, $\delta_\text{ex/abs}$, and type formation context, $\Delta_\text{ex/abs}$, from $\memD_\text{ex/abs}$, again equivalent to $\delta_\text{conf/abs}$ and $\Delta_\text{conf/abs}$ from Sec.
 \ref{sec:selective-type-translation-abstraction}.
@@
 
-Applying the substitutions $\gamma_\text{ex/abs}$ and $\delta_\text{ex/abs}$ in the conclusion of the rule, we arrive at the actual term translation $\iota_\text{ex}$.
+Applying the substitutions $\gamma_\text{ex/abs}$ and $\delta_\text{ex/abs}$ in the conclusion of the rule, we arrive at the actual term translation $\iota_\text{ex}$, defined previously.
+Note that $\iota_\text{ex}$ has type $\tau_\text{paper}$ under the translation of $\Upsilon_\text{ex}$, i.e.
+$\vdash \Upsilon_\text{ex} \leadsto \Gamma_\text{ex}$ where $\Gamma_\text{ex} := title : \keyw{str}$.
 
@@
 The term index, rather than the tycon context, directly specifies the static function that maps the arguments to a translation.
-In all other respects, the rule is analagous.
-It is used as a technical device in our proof of conservativity in Sec.
+In all other respects, it is analagous.
+It is used as a technical device in Sec.
 \ref{metatheory}.
@@
 
-Desugarings include an explicit form, $e_\text{targ}.\opname{op}\langle\sttmidx\rangle(\es)$ (and variants of where the term index or the arguments are omitted), projection syntax for use by record-like types, $e_\text{targ}\opname{\#}\conclbl{lbl}$, which desugars to $\etarg{\opname{prj}}{\conclbl{lbl}}{e_\text{targ}}{\cdot}$, and concatenative syntax $e_\text{targ} \cdot e_\text{arg}$, which desugars to $\etarg{\opname{conc}}{\striv}{e_\text{targ}}{e_\text{arg}}$.
+Concrete desugarings include $e_\text{targ}.\opname{op}\langle\sttmidx\rangle(\es)$ (and variants where the term index or arguments are omitted), projection syntax for use by record-like types, $e_\text{targ}\opname{\#}\conclbl{lbl}$, which desugars to $\etarg{\opname{prj}}{\conclbl{lbl}}{e_\text{targ}}{\cdot}$, and $e_\text{targ} \cdot e_\text{arg}$, which desugars to $\etarg{\opname{conc}}{\striv}{e_\text{targ}}{e_\text{arg}}$.
 We show other desugarings, including case analysis, in the supplement.
@@
 The second premise extracts the tycon definition for $\tc$ from the tycon context.
-The third extracts the \emph{operator index kind} from its opcon signature, and the fourth checks the term index against this kind.
-For example, we may associate the following opcon signatures with the tycons $\tcvar{rstr}$ and $\tcvar{lprod}$:\[\small
-\begin{array}{lcl}
-\chi_\text{rstr} & := & \keyw{intro}[\keyw{Str}], \opsig{\opname{conc}}{\kunit}, \opsig{\opname{case}}{\klist{\keyw{StrPattern}}},  \\
-& & \opsig{\opname{coerce}}{\krx}, \opsig{\opname{check}}{\krx}, \opsig{\opname{replace}}{\krx}\\
-\chi_\text{lprod} & := & \keyw{intro}[\klist{\klbl}], \opsig{\opname{prj}}{\klbl}, \opsig{\opname{conc}}{\kunit}, \opsig{\opname{drop}}{\klist{\klbl}}
-\end{array}
-\]
-
+The third extracts the \emph{operator index kind} from its opcon signature, and the fourth checks the term index against it.
+% :\[\small
+% \begin{array}{lcl}
+% \chi_\text{rstr} & := & \keyw{intro}[\keyw{Str}], \opsig{\opname{conc}}{\kunit}, \opsig{\opname{case}}{\klist{\keyw{StrPattern}}},  \\
+% & & \opsig{\opname{coerce}}{\krx}, \opsig{\opname{check}}{\krx}, \opsig{\opname{replace}}{\krx}\\
+% \chi_\text{lprod} & := & \keyw{intro}[\klist{\klbl}], \opsig{\opname{prj}}{\klbl}, \opsig{\opname{conc}}{\kunit}, \opsig{\opname{drop}}{\klist{\klbl}}
+% \end{array}
+% \]
+
+Figure \ref{fig:example-tycons} showed portions of the opcon signatures of $\tcvar{rstr}$ and $\tcvar{lprod}$.
 The opcons associated with $\tcvar{rstr}$ are taken directly from Fulton et al.'s specification of regular string types \cite{sanitation-psp14}, with the exception of $\opname{case}$, which generalizes case analysis as defined there to arbitrary string patterns, which we discuss in the supplement.
 The opcons associated with $\tcvar{lprod}$ are also straightforward: $\opname{prj}$ projects out the row with the provided label, $\opname{conc}$ concatenates two labeled products (updating common rows with the value from the right argument), and $\opname{drop}$ creates a new labeled product from the target with some rows dropped.
-Note that both regular strings and labeled products can define concatenation without conflict.
+Note that both $\tcvar{rstr}$ and $\tcvar{lprod}$ can define concatenation without conflict.
 %Targeted operations use no new mechanisms, so we only show regular string concatenation; others are in the supplement.
@@
 The rule (ocstruct-targ) in Figure \ref{ocstruct} ensures that no tycon defines an opcon twice and that the opcon definitions are well-kinded.
-For example, $\omega_\text{rstr/targops}$ defines:
+For example, $\stx{rstr/conc}=$
 \[\small
@@
 \section{Metatheory}\label{metatheory}
-We will now state the key metatheoretic properties of @$\lambda$. The full proofs are in the supplement.
+\noindent We will now state the key metatheoretic properties of @$\lambda$. The full proofs are in the supplement.
 
@@
 Variables in $\Upsilon$ can  occur in arguments (e.g.
-$title$ in the previous example), but the translations of the arguments only appear \emph{after} the substitution $\gamma$ has been applied.
+$title$ in the earlier example), but the translations of the arguments only appear \emph{after} the substitution $\gamma$ has been applied.
 We assume that substitution is capture-avoiding in the usual manner.
@@
 \end{tyconinvariant}
-\begin{proof-sketch} The proof is not unusually difficult because we have fixed the tycon context $\Phi_\text{rstr}$, so we can essentially treat the calculus like a type-directed compiler for a calculus with only two tycons, $\rightharpoonup$ and $\tcvar{rstr}$.
+\begin{proof-sketch} The proof is not unusually difficult because we have fixed the tycon context $\Phi_\text{rstr}$, so we can essentially treat the calculus like a type-directed compiler for a calculus with only two tycons, $\rightharpoonup$ and $\tcvar{rstr}$, plus one ``other'' one.
 Such a calculus and compiler specification was given in \cite{sanitation-psp14}, so we must simply show that the opcon definitions in $\tcvar{rstr}$ adequately capture these specifications using standard techniques for the SL, a simply-typed functional language \cite{conf/pldi/Chlipala07}.
 The only ``twist'' is that the rule (syn-targ-other) can synthesize a regular string type.
-But if so, the abstracted translation must have type $\tau_\text{abs}=\alpha$.
-Thus, invariants cannot be violated by relational parametricity in the IL (i.e.
+But if so, $\iota_\text{abs}$ will be checked against $\tau_\text{abs}=\alpha$.
+Thus, the invariants cannot be violated by direct application of relational parametricity in the IL (i.e.
 a ``free theorem'') \cite{WadlerThms}.
@@
 
-The reason why (syn-targ-other) is never a problem in proving a tycon invariant, i.e.
-\emph{translation independence} of tycons, turns out to be the same reason extending the tycon context conserves all tycon invariants.
+The reason why (syn-targ-other) is never a problem in proving a tycon invariant -- \emph{translation independence} of tycons -- turns out to be the same reason extending the tycon context conserves all tycon invariants.
 A newly introduced tycon defining a targeted operator that synthesizes a regular string type, e.g.
@@
 %Type abstraction is the basis for conservatively composing type system fragments, just like it is the basis for composing ML-style modules.
+
  We can state this more generally:
@@
 The proof maps every well-typed term under $\Phi'$ to a well-typed term under $\Phi$ with the same translation, and if the term has a type constructed by a tycon in $\Phi$, e.g.
-$\tc$, the mapping still has a type constructed by that tycon with the same type translation.
-If $P(\sttyidx, \iota)$ is preserved under the mapping over $\sttyidx$ then we can simply invoke the assumed tycon invariant.
+$\tc$, the new term has a type constructed by that tycon with the same type translation, and only a slightly different type index.
+In particular, the mapping's effect on static terms is to replace all types constructed by $\tc'$ with a unique type constructed by $\keyw{other}[m]$ (for some suitable $m$ not already used).
+If $P(\sttyidx, \iota)$ is preserved under this transformation then we can simply invoke the existing proof of the tycon invariant.
 We call such propositions \emph{modular}.
-Non-modular propositions are uninteresting because the mapping's effect on static terms is only to replace all types constructed by $\tc'$ with a unique type constructed by $\keyw{other}[m]$, for $m$ uniquely associated with $\tc'$.
-Any propositions decidable by the SL are modular.
+Non-modular propositions are uninteresting because they have knowledge of tycons ``from the future''.
+
 On external terms, the mapping replaces all intro and targeted terms associated with $\tc'$ with an equivalent one that passes through the rules (ana-intro-other) and (syn-targ-other) by partially applying the intro and targeted opcon definitions to generate the term indices.
 Typing, kinding, static normalization and selective translation abstraction are preserved under the mapping, defined inductively in the supplement.
+Note that for this reason all propositions decidable by the SL are modular.
 \end{proof-sketch}
@@
 Such a system could be seen in some ways as a degenerate mode of use of our work: we further cover the cases when new  operators are needed. 
-For example, the $\tcvar{lprod}$ opcons use labels and types, which exist only statically; they are not refinements of binary products because no row projection operator exists.
+For example, the $\tcvar{lprod}$ opcons use labels and types, which exist only statically.
+Thus, labeled tuples cannot be seen as refinements of nested pairs because the row projection operators don't exist.
 %Labeled products are not refinements of binary products because the label projection operator is not defined for binary products.%Regular string types could also support an operation that the internal string type would not (e.g.
@@
 canonical forms) \cite{conf/popl/DelawareOS13}.
-Both there and in work by Chlipala  \cite{Chlipala10}, the suggestion is to use  proof automation, but this is a heuristic approach.
-
-Fragment providers need not develop mechanized proofs with our approach.
-Under a fixed tycon context, the calculus can be reasoned about like a very small type-directed compiler \cite{tarditi+:til-OLD,conf/pldi/Chlipala07}.
-Errors in an opcon definition lead to failure when typechecking external terms that use them.
+The authors, along with Chlipala  \cite{Chlipala10}, suggest  proof automation. This is fundamentally a heuristic approach.
+
+A strength of @$\lambda$ is that fragment providers need not provide the semantics with mechanized proofs to modularly reason about language extension.
+Instead, under a fixed tycon context, the calculus can be reasoned about like a very small type-directed compiler \cite{tarditi+:til-OLD,conf/pldi/Chlipala07}.
+Errors in reasoning can only lead to failure at typechecking time, a form of \emph{translation validation} \cite{Pnueli-Siegel-Singerman98}.
 Incorrect opcon definitions (relative to a specification, e.g.
-\cite{fulton-thesis} for regular strings) can only weaken expected invariants at that tycon, like incorrectly implemented modules in ML, so tycons can reasonably be tested in real code without worrying about non-local impact.
-The simplicity of opcons also suggests educational applications (like TinkerType \cite{LevinPierce99}).
-To eliminate even these localized failure modes for ``reliability-critical'' tycons, we are working on \emph{optional} proof mechanization in the SL (by making it dependently kinded), and on mechanizing the metatheory of @$\lambda$ itself.%Every combination of fragments is a new dialect, and must be reasoned about monolithically.
+\cite{sanitation-psp14} for regular strings) can at worst weaken expected invariants at that tycon, like incorrectly implemented modules in ML.
+Thus, modular tycons can reasonably be tested ``in the field'' without concern about the reliability of the system as a whole.
+The simplicity of writing extensions directly in a functional language also suggests educational applications (following TinkerType \cite{LevinPierce99}).
+To eliminate even these localized failure modes for ``reliability-critical'' tycons, we plan to introduce \emph{optional} proof mechanization into the SL (by basing it on a dependently typed language like Coq) in subsequent work.
+%We also hope to mechanize the metatheory of @$\lambda$ itself in future work.%Every combination of fragments is a new dialect, and must be reasoned about monolithically.
 %Efforts to reason modularly areeed, several problems can come up when fragments are combined, so .
@@
 
-Type abstraction, encouragingly, also underlies modular reasoning in ML-like languages \cite{pfpl,harper1997programming,conf/oopsla/AminRO14} and languages with other forms of ADTs \cite{liskov1974programming}.
-Indeed, proofs of tycon invariants can rely on the same parametricity theorem \cite{WadlerThms}.
+Type abstraction, encouragingly, also underlies modular reasoning in ML-like languages \cite{pfpl,harper1997programming} and languages with other forms of ADTs \cite{liskov1974programming} like Scala \cite{conf/oopsla/AminRO14}.
+Indeed, proofs of tycon invariants can rely on existing parametricity theorems \cite{WadlerThms}.
 Our work is technically reminiscent of  work on  translating modules to System $\mathbf{F}_\omega$ \cite{conf/tldi/RossbergRD10}.
-Unlike in module systems, type translations (analagous to the choice of representation for an abstract type) are \emph{statically computed} based on a type index, rather than declared directly.
-Moreover, there can be arbitrarily many operators because they arise by providing a term index to an opcon, and their  semantics can be complex because a static function computes types and translations.
+Unlike in module systems, type translations (analagous to the choice of representation for an abstract type) are statically \emph{computed} based on a type index, rather than statically \emph{declared}.
+Moreover, there can be arbitrarily many operators because they arise by providing a term index to an opcon, and their  semantics can be complex because a static function computes the types and translations that arise.
 In contrast, modules and ADTs can only specify a fixed number of operations, and each must have function type.
 Note also that these are not competing mechanisms: we did not specify quantification over external types here for simplicity, but we conjecture that it is complementary and thus @$\lambda$ could serve as the core of a language with an ML-style module system.
-Another related direction is \emph{tycon functors}, which would abstract over tycons with the same signature to support  different cost models.
+Another related direction is \emph{tycon functors}, which would abstract over tycons with the same signature to support a tunable cost semantics.
 % This permits us tocode targeting a separate IL, and permit operations whose types cannot simply be written as functions.
@@
 
-One limitation of our approach is that it supports only  fragments with the standard ``shape'' of typing judgement.
-Fragments that require new forms of contexts (e.g.
-symbol contexts \cite{pfpl}), for example, cannot be defined.
+A limitation of our approach is that it supports only  fragments with the standard ``shape'' of typing judgement.
+Fragments that require new forms of scoped contexts (e.g.
+symbol contexts \cite{pfpl}) or unscoped declarations cannot presently be defined.
 Relatedly, the language controls variable binding, so, for example, linear type systems, cannot be defined.
 Another limitation is that opcons cannot directly invoke one another (e.g.
-a \opname{len} opcon on regular strings could not construct a natural number).
-We conjecture that these are not fundamental limitations and expect @$\lambda$ to serve as a minimal foundation for future efforts that increase  expressiveness while maintaining the fundamental guarantees established  here.
+a \opname{len} opcon on regular strings could not construct a natural number; see supplement for a workaround, however).
+We conjecture that these are not fundamental limitations and expect @$\lambda$ to serve as a minimal foundation for future efforts that increase  expressiveness while maintaining the strong guarantees, like type safety and conservativity, established  here.
 %(tycons could define new contexts which are threaded opaquely through ``outside'' operations).

### 2014-11-04 att-pldi15/att-pldi15.tex
updated

@@
 
-In this paper, we begin from first principles by specifying a language, @$\lambda$, in the style of many full-scale languages: as a bidirectionally typed translation semantics.
-Only the $\rightarrow$ type constructor (tycon) is built in; all other external tycons (we show constrained strings and a variant on records) are defined by extending a \emph{tycon context}.
-Each tycon defines the semantics of its  term-level operators (e.g.
-record projection) using a static language where types and translations are values.
-The semantics includes a simple check that leads to powerful metatheoretic guarantees, notably \emph{type safety} and \emph{conservativity}: that  \emph{tycon invariants} will always be conserved under extension.
+In this paper, we begin from first principles by specifying a language, @$\lambda$, in a style that follows many full-scale languages: as a bidirectionally typed translation semantics.
+Only the $\rightarrow$ type constructor (tycon) is built in; all other external tycons (we show constrained strings and a variation on record types) are defined by extending a \emph{tycon context}.
+Each tycon defines the semantics of its associated term-level \emph{opcons} (e.g.
+row projection) using a static language where types and translations are values.
+The semantics guarantees \emph{type safety} and, unlike other approaches, \emph{conservativity}: that all \emph{tycon-specific invariants} that can be established in a ``closed world'' will be conserved in the ``open world''.
 Type system providers need not provide mechanized proofs.
-Instead, the check enforces a \emph{translation independence} property via type abstraction, the same principle underlying ML-style module systems.
+Instead, these follow from a form of translation validation that uses type abstraction to check that tycons maintain \emph{translation independence}, analagous to the representation independence maintained by ML-style module systems.
 \end{abstract}
@@
 Universality properties (e.g.
-``Turing-completeness'') are often invoked to guarantee that an embedding that preserves a desirable fragment's dynamics can be constructed, but an isomorphic embedding must also preserve the static semantics and, if defined, performance bounds specified using a cost semantics.
+``Turing completeness'') are often enough to guarantee that an embedding that preserves a desirable fragment's dynamic semantics can be constructed, but an isomorphic embedding must also preserve the static semantics and, if defined, performance bounds specified using a cost semantics.
 This is not always possible.
-Embeddings are also sometimes too \emph{complex}, as measured by the cost of the extralinguistic computations that are needed to map in and out of the embedding and, if these  must be performed mentally by programmers, considering various human factors.
+For example, in $\mathcal{L}\{{\rightharpoonup}\,{\forall}\,{\mu}\,{\keyw{1}}\,{\times}\,{+}\}$, it is impossible to introduce record types as a library because you need the row projection operators, in ML written $\opname{\#}\conclbl{lbl}$, one for each of the infinite set of row labels $\conclbl{lbl}$ ($\opname{\#}$ itself is an \emph{operator constructor}).
+%You can only use $\mathcal{L}\{{\rightharpoonup}\,{\forall}\,{\mu}\,{\keyw{1}}\,{\times}\,{+}\}$ as a \emph{target of compilation} for a language with record types (they can be implemented using nested binary products,  lists, or other types).
+%(and attempting to simulate records via a hash table over strings plus a refinement type system to check that the lookup operation will not fail will not satisfy a cost semantics prescribing $\mathcal{O}(1)$ lookup; see Sec.
+\ref{prior-work}).
+
+%Embeddings are also sometimes too \emph{complex}, as measured by the cost of the extralinguistic computations that are needed to map in and out of the embedding and, if these  must be performed mentally by programmers, considering various human factors.
 %We will discuss specific examples below.
@@
 %Indeed, the proliferation of language dialects constructed for this reason might be taken as an evidence that the core language is not as ``general'', when considered comprehensively, as might be hoped.
-Each time a fragment must be exposed directly, rather than as a library, a new \emph{dialect} of the language must be constructed.
-Within the ML lineage, for example, dialects that go beyond ``core ML'' abound:
+Each time a fragment must be introduced directly, rather than as a library, a new \emph{dialect} of the language arises.
+Within the ML lineage, dialects that go beyond ``core ML'' abound:%, showing that this is not an isolated example:
 %Reynolds, in a remark that recalls the ``Turing tarpit'' of Perlis \cite{Perl82a}, summarizes the issue \cite{Reynolds94anintroduction}: 
@@
 \begin{enumerate}
-\vspace{-5px}
+\vspace{-7px}
 %\compresslist
@@
 \textbf{General Purpose Fragments:} 
-A number of variations on product types, for example, have been introduced in dialects: 
+Many variations on product types, for example, exist: 
 $n$-ary tuples, 
@@
 records with ``methods'' (i.e. pure objects \cite{TSLs}). 
- Sum-like types are also exposed in various ways: 
-finite datatypes, 
-open datatypes \cite{conf/ppdp/LohH06}, 
-hierarchically open datatypes \cite{journals/toplas/MillsteinBC04}, 
+ Sum-like types are also exposed variously: 
+standard datatypes, 
+open datatypes \cite{conf/ppdp/LohH06,journals/toplas/MillsteinBC04}, 
 polymorphic variants \cite{ocaml-manual} and 
-ML-style exception types. Combinations of these manifest themselves as class-based object systems \cite{ocaml-manual}.
+exception types \cite{harper1997programming}. Combinations of these manifest themselves as class-based object systems \cite{ocaml-manual}.
 
 \item
-\textbf{Specialized Fragments:} Fragments that track specialized static invariants to provide stronger correctness or security guarantees, manage unwieldy lower-level abstractions and run-time systems or control cost are also frequently introduced in dialects, e.g.
-for data parallelism  \cite{chakravarty2007data}, distributed programming \cite{Murphy:2007:TDP:1793574.1793585}, reactive programming \cite{mandel2005reactiveml}, authenticated data structures \cite{Miller:2014:ADS:2535838.2535851}, databases \cite{Ohori:2011:MSM:2034773.2034815},  units of measure \cite{conf/cefp/Kennedy09} and regular string sanitation \cite{sanitation-psp14}.
+\textbf{Specialized Fragments:} Fragments that track specialized static invariants are also frequently introduced in dialects, e.g.
+for distributed programming \cite{Murphy:2007:TDP:1793574.1793585}, reactive programming \cite{mandel2005reactiveml}, authenticated data structures \cite{Miller:2014:ADS:2535838.2535851}, databases \cite{Ohori:2011:MSM:2034773.2034815},  units of measure \cite{conf/cefp/Kennedy09} and regular string sanitation \cite{sanitation-psp14}, amongst many other examples.
 %\verb|sprintf| in the OCaml dialect statically distinguishes format strings from strings.% All of these are implemented as dialects of existing languages, presumably because a strong encoding was not feasible.
@@
 a safe FFI, item 3 above, is needed between every pair of dialects).
-And as just mentioned, composing dialects is non-trivial.
-
-These problems do not arise for a fragment that can be expressed as an isomorphic embedding (i.e.
+%And as just mentioned, composing dialects is non-trivial.
+
+These problems do not arise for fragment expressed as an isomorphic embedding (i.e.
 as a library) because modern \emph{module systems} can enforce abstraction barriers that ensure that the isomorphism needs only to be established in the ``closed world'' of the module.
-For example, a module defining sets in ML can hold the representation of sets abstract, ensuring that any invariants maintained by the functions in the module (e.g.
-uniqueness, if using a list representation) will hold no matter which other modules are in use by a client \cite{harper1997programming}.
-Other languages provide similar facilities, e.g.
-Scala's abstract type members \cite{conf/oopsla/AminRO14}.
+For example, a module defining sets in ML can hold the representation of sets abstract, ensuring that any invariants necessary to maintain the isomorphism need only be maintained by the functions in the module (e.g.
+uniqueness, if using a list representation).
+These then continue to hold no matter which other modules are in use by a client \cite{harper1997programming}.
+Other languages provide similar forms of \emph{abstract data types} to localize reasoning \cite{liskov1974programming}.
 %Mechanisms that can help decrease the complexity of an embedding without violating abstraction barriers are thus valuable, and we will lead into our work in Sec.
@@
 
-When library-based embeddings are not possible, as in the examples above, mechanisms are needed that make it possible to define and  reason in a similarly modular manner about  direct extensions to the semantics of a language.
-Such a mechanism could ultimately be integrated directly into the language, blurring the distinction between fragments and libraries and decreasing the need for new dialects.% Importing fragments that introduce new semantics   would be as easy and reasonable as importing a new module is in ML today.
+Mechanisms are needed that make it possible to define and  reason in a similarly modular, localized manner about  direct extensions to the semantics of a language.
+For example, if a language is extended with \emph{regular string types} as described in \cite{sanitation-psp14}, all terms having a regular string type like $\sty{\tcvar{rstr}}{\concrx{[A-Z]+ \digit\digit\digit\digit}}$ should continue to behave as strings in the regular language indicated no matter which other extensions are in use by a client.
+%Such a mechanism could ultimately be integrated directly into the language, blurring the distinction between fragments and libraries and decreasing the need for new dialects.% Importing fragments that introduce new semantics   would be as easy and reasonable as importing a new module is in ML today.
 %In the limit, the community could rely on modularly mechanized metatheory and compiler correctness results.% rather than requiring heroic efforts from individual research groups that consider an entire language at once.
@@
 Rather than building in a monolithic set of external type constructors, however, the  semantics are indexed by a \emph{tycon context}.
-Each tycon defines the semantics of its operators via functions written in a \emph{static language} (SL) where types and translations are values.
-%, which provides a form of type-level computation.
+Each tycon, e.g.
+$\tcvar{lprod}$ defining labeled products or $\tcvar{rstr}$ defining regular strings, determines the semantics of its associated opcons (e.g.
+$\opname{\#}$ for row projection, or $\opname{conc}$ for concatenation) via \emph{static functions}, i.e.
+functions written in a \emph{static language} (SL), where types and translations are values.
+
+All translation invariants maintained by these opcons in any ``closed world'', e.g.
+the regular string invariant just mentioned, are guaranteed to be maintained in any further extended tycon context, i.e.
+in the ``open world'', due to a simple static check that maintains \emph{translation independence} between tycons using type abstraction in the IL, the same fundamental mechanism underlying representation independence in ML-style module systems.
+As in ML, mechanized specifications and proofs are not needed.%, which provides a form of type-level computation.
 %The authors demonstrate the expressive power of this technique primarily with an implementation and a number of examples of fragments as libraries, and their core calculus .% (unlike systems that treat the type system as a ``bag of rules'', where non-determinism can arise).
 
-We will begin by giving an overview of the organization  and main judgements of @$\lambda$ in Sec.
-\ref{atlam}, then discuss how types are constructed and introduce our two main examples, one defining labeled product types with a functional update operator, and the other regular string types, based on a recent core calculus style specification \cite{sanitation-psp14}, in Sec.
-\ref{types}.
-We describe how tycons control the semantics of their term-level operator constructors (opcons) in Sec.
+% We will begin by giving an overview of  @$\lambda$ in Sec.
+\ref{atlam}, introducing our two main examples, one defining labeled product types (i.e.
+record types maintaining an ordering for simplicity) with a functional update operator, and the other regular string types, based on a recent core calculus style specification \cite{sanitation-psp14}.
+In Sec.
+\ref{types}, we detail how tycons are defined and how types arise from tycons.
+We then describe how tycons control the semantics of their term-level operator constructors (opcons) in Sec.
 \ref{external-terms}.
@@
 in the ``open world'').
-Interestingly, type system providers need not provide mechanized proofs to maintain these guarantees.
+Interestingly, type system providers need not provide mechanized proofs to maintain such guarantees.
 Instead, the approach we take relies on type abstraction in the internal language.
@@
 Programmers interface with @$\lambda$ by writing \emph{external terms}, $e$.
-The abstract syntax of external terms is shown in Figure \ref{syntax-EL} (we will give various concrete desugarings in Sec.
-\ref{external-terms}).
+The abstract syntax of external terms is shown in Figure \ref{syntax-EL} and we will introduce various concrete desugarings in Sec.
+\ref{external-terms}.
 %We will describe useful syntactic desugarings atop this syntax as we go on and review recent techniques that permit modularly introducing  desugarings like these in Sec.
@@
 Our main focus here is on semantic extensions.
-The static and dynamic semantics are specified simultaneously as a \emph{bidirectionally typed translation semantics}, i.e.
-the key judgements take the form:\[\esynX{e}{\st\moutput}{\iota\moutput} ~~~~~\text{and}~~~~~ \eanaX{e}{\st}{\iota\moutput}\]
+The semantics are specified as a \emph{bidirectionally typed translation semantics}, i.e.
+the key judgements have the form:\[\esynX{e}{\st\moutput}{\iota\moutput} ~~~~~\text{and}~~~~~ \eanaX{e}{\st}{\iota\moutput}\]
 \noindent
 These are pronounced ``$e$ (synthesizes / analyzes against) type $\sigma$ and has  translation $\iota$ under typing context $\Upsilon$ and tycon context $\Phi$''.
-Note that our specifications in this paper are intended to be algorithmic: we indicate ``outputs'' when introducing judgement forms by \emph{mode annotations}, $\moutput$; these are not part of the judgement's syntax.
-In particular, note that the type is an ``output'' only for the synthetic judgement.
-
-This basic separation of the EL and IL is commonly used for full-scale language specifications, e.g.
-the Harper-Stone semantics for Standard ML \cite{Harper00atype-theoretic}.
+Note that our specifications in this paper are intended to be algorithmic: we indicate ``outputs'' when introducing judgement forms by \emph{mode annotations}, $\moutput$.
+Note that the type is an ``output'' only for the synthetic judgement.
+
+This separation of the EL from the IL is commonly used for full-scale language specifications, e.g.
+the Harper-Stone semantics for Standard ML \cite{Harper00atype-theoretic} or the semantics of Wyvern \cite{TSLs}.
 The internal language is purposely kept small, e.g.
@@
 In @$\lambda$, the EL builds in only function types.
-All other external constructs are defined in the \emph{tycon context}, described starting in the Sec.
+All other external constructs are defined in the \emph{tycon context}, $\Phi$, described starting in the Sec.
 \ref{types}.
@@
 The first is once again to justify the practicality of our approach: local type inference is increasingly being used in modern languages (e.g.
-Scala \cite{OdeZenZen01}) because it eliminates the need for type annotations in many situations while remaining decidable in more situations than whole-function type inference and providing what are widely perceived to be higher quality error messages \cite{journals/jfp/JonesVWS07}.
+Scala \cite{OdeZenZen01}) because it eliminates the need for type annotations in many situations while being simpler and decidable in more situations than whole-function type inference and providing what are widely perceived to be higher quality error messages \cite{journals/jfp/JonesVWS07}.
 Secondly, it  gives us a clean way to reuse the generalized introductory form, $\eintro{\st}{\es}$, and its associated desugarings, at many types \cite{TSLs}.
@@
 \ref{sec:type-translations}.
-For this reason, this specification style may also be compared to specifications for the first stage of a type-directed compiler, e.g.
+For example, regular string types will translate to internal type abbreviated $\keyw{str}$, and labeled product types will translate to nested binary product types, though we will emphasize that there are other valid choices.
+
+This specification style may thus also be compared to specifications for the first stage of a type-directed compiler, e.g.
 the TIL compiler for Standard ML \cite{tarditi+:til-OLD}, here lifted ``one level up'' into the semantics of the language itself.
@@
 
-External typing contexts $\Upsilon$ map variables to types, so we also need the judgement $\vdash_\Phi \Upsilon \leadsto \Gamma$ to construct a $\Gamma$ that maps the same variables to corresponding type translations.
+External typing contexts $\Upsilon$ map variables to types, so we also need the judgement $\vdash_\Phi \Upsilon \leadsto \Gamma$.
 
@@
 \end{figure}
-The workhorse of @$\lambda$ is the \emph{static language}, which itself forms a typed lambda calculus where 
+The main novelty of @$\lambda$ is the \emph{static language}, which itself forms a typed lambda calculus where we use the word 
 %External terms are classified by (external) \emph{types}. 
-\emph{kinds}, $\kappa$, classify \emph{static terms}, $\sigma$.
+\emph{kind}, $\kappa$, to refer to the ``types'' of \emph{static terms}, $\sigma$.
 The syntax of the SL is given in Figure \ref{syntax-SL}.
-The portion of the SL covered by the first row of kinds and first row of static terms, some of which are elided for concision, forms an entirely standard  functional programming language consisting of total functions, universal quantification over kinds, inductive kinds, and products and sums.
-The reader can consider these as forming a total subset of ML or a simply-typed subset of Coq and we will assume standard conveniences from such languages (e.g.
-let bindings, pattern matching and inference of type parameters when applying polymorphic functions) are available in examples for concision.
-The semantics we assume also directly follows \cite{pfpl}, so we also omit the details of this core here.
-Only three new  kinds are needed for the SL to serve its role as the language used to   control typing and translation of the EL: $\kty$ (Sec.
-\ref{types}), $\kity$ (Sec.
-\ref{sec:type-translations}) and $\kitm$ (Sec.
+The portion of the SL covered by the first row of kinds and static terms, some of which are elided for concision, forms an entirely standard total functional programming language consisting of total functions, universal quantification over kinds, inductive kinds, and products and sums  \cite{pfpl}.
+The reader can consider these as forming a total subset of ML \cite{harper1997programming} or a simply-typed subset of Coq \todo{citation}and we will assume standard conveniences from such languages (e.g.
+let bindings, pattern matching and inference of type parameters when applying polymorphic functions) in examples for concision.
+Only three new  kinds are needed for the SL to serve its role as the language used to   control typing and translation of the EL: $\kty$, classifying types (Sec.
+\ref{types}), $\kity$, classifying \emph{translational internal types} (Sec.
+\ref{sec:type-translations}) and $\kitm$, classifying \emph{translational internal terms} (Sec.
 \ref{sec:introop}).
@@
 All such contexts in @$\lambda$ are identified up to exchange and contraction and obey weakening \cite{pfpl}.
-The natural number $n$ is used as a technical device in our semantics to ensure that the forms shown as being indexed by $n$ in the syntax only arise in a controlled manner internally to prevent ``out of bounds'' issues, as we will discuss; they would have no corresponding concrete syntax so $n$ can be assumed $0$ in user-defined terms.
-
-The dynamic semantics of static terms is defined as a structural operational semantics by a stepping judgement $\sstep{\st}{\argEnv}{\st\moutput}$, a value judgement $\sval{\st}{\argEnv}$ and an error judgement $\serr{\st}{\argEnv}$.
-Here, $\argEnv$ ranges over \emph{argument environments}, which we will  return to when considering opcons in Sec.
+The natural number $n$ is used as a technical device in our semantics to ensure that the forms shown as being indexed by $n$ arise in a controlled manner to prevent ``out of bounds'' issues, as we will discuss; they would have no corresponding concrete syntax so $n$ can be assumed $0$ in user-defined terms.
+
+The dynamic semantics of static terms is defined as a structural operational semantics by a stepping judgement $\sstep{\st}{\argEnv}{\st\moutput}$, a value judgement $\sval{\st}{\argEnv}$ and an error raised judgement $\serr{\st}{\argEnv}$.
+$\argEnv$ ranges over \emph{argument environments}, which we will  return to when considering opcons in Sec.
 \ref{external-terms}.
@@
 We omit error propagation rules.%We will refer to the relevant rules as we proceed.
+
+\section{Types}\label{types}
+
+\begin{figure*}\begin{mathpar}
+\small
+\inferrule[k-parr]{
+    \sofkX{\st}{\kprod{\kty}{\kty}}
+}{
+    \sofkX{\sty{\rightharpoonup}{\st}}{\kty}
+}
+
+\inferrule[k-tc]{
+    \tcdef{\tc}{\tcsig{\ktyidx}{\chi}}{\theta} \in \Phi\\
+    \sofkX{\st}{\ktyidx}
+}{
+    \sofkX{\sty{\tc}{\st}}{\kty}
+}
+
+\inferrule[k-otherty]{
+    \sofkX{\sttyidx}{\kappa \times \kity}
+}{
+    \sofkX{\sty{\keyw{other}[m; \kappa]}{\sttyidx}}{\kty}
+}\vspace{-10px}
+\end{mathpar}
+\caption{Kinding rules for types, which take the form $\sty{c}{\sttyidx}$ where $c$ is a tycon and $\sttyidx$ is the type index.}
+\label{fig:types}\vspace{-10px}
+\end{figure*}
 \begin{figure}[t]
@@
 $\begin{array}{lrcl}
-\textbf{tycons} & c & ::= & \rightharpoonup ~|~ \tc ~|~ \keyw{other}[m]\\
+\textbf{tycons} & c & ::= & \rightharpoonup ~|~ \tc ~|~ \keyw{other}[m; \kappa]\\
 \textbf{tycon contexts} & \Phi & ::= & \cdot ~|~ \Phi, \tcdef{\tc}{\psi}{\theta}\\
@@
 \end{array}$
-\caption{Syntax of tycons. Metavariables $\tc$ and $\opname{op}$ range over user-defined tycon and opcon names, respectively, and $m$ ranges over natural numbers.}
+\caption{Syntax of tycons. Metavariables $\tc$ and $\opname{op}$ range over extension tycon and opcon names, respectively, and $m$ ranges over natural numbers.}
 \label{syntax-TC}\vspace{-8px}
 \end{figure}
-\begin{figure*}\begin{mathpar}
-\small
-\inferrule[k-parr]{
-    \sofkX{\st}{\kprod{\kty}{\kty}}
-}{
-    \sofkX{\sty{\rightharpoonup}{\st}}{\kty}
-}
-
-\inferrule[k-ty]{
-    \tcdef{\tc}{\tcsig{\ktyidx}{\chi}}{\theta} \in \Phi\\
-    \sofkX{\st}{\ktyidx}
-}{
-    \sofkX{\sty{\tc}{\st}}{\kty}
-}
-
-\inferrule[k-otherty]{
-    \sofkX{\sttyidx}{\keyw{Nat} \times \kity}
-}{
-    \sofkX{\sty{\keyw{other}[m]}{\sttyidx}}{\kty}
-}\vspace{-10px}
-\end{mathpar}
-\caption{Kinding rules for types, which take the form $\sty{c}{\sttyidx}$ where $c$ is a tycon and $\sttyidx$ is the type index.}
-\label{fig:types}\vspace{-10px}
-\end{figure*}
-
-\section{Types}\label{types}
-
+
+\noindent
 External types, or simply \emph{types}, are static values of kind $\kty$.
+We write $\istype{\st}{\Phi}$  iff $\sofkn{\emptyset}{\emptyset}{\Phi}{0}{\st}{\kty}$ and $\sval{\st}{\cdot;\emptyset;\Phi}$.
 The introductory form for kind $\kty$ is $\sty{c}{\st}$, where $c$ is a \emph{tycon} and $\st$ is the \emph{type index}.
-The syntax  for tycons given in Figure \ref{syntax-TC} specifies that $c$ is either the built-in tycon governing partial functions, $\rightharpoonup$, a user-defined tycon name written in small caps, $\tc$, or an ``other'' tycon, $\keyw{other}[m]$, indexed by a natural number $m$ to allow arbitrarily many such tycons.
-The kinding rules governing the form $\sty{c}{\st}$ are shown in Figure \ref{fig:types}.
-The dynamics are simple (see supplement): the index is eagerly normalized and errors propagate.
-We write $\istype{\st}{\Phi}$  iff $\sofkn{\emptyset}{\emptyset}{\Phi}{0}{\st}{\kty}$ and $\sval{\st}{\cdot;\emptyset;\Phi}$.
-
-The rule \rulename{k-parr} specifies that the type index of partial function types must be a pair of types.
+The dynamics are simple: the index is eagerly normalized and errors propagate  (see supplement for the complete set of rules in this paper).
+
+The syntax  for tycons given in Figure \ref{syntax-TC} specifies that $c$ can have one of three forms, so there are three corresponding kinding rules for types, shown in Figure \ref{fig:types}.
+
+\paragraph{Arrow Types} The rule \rulename{k-parr} specifies that the type index of partial function types must be a pair of types.
 We thus say that $\rightharpoonup$ has \emph{index kind} $\kty \times \kty$.
-To recover a conventional syntax, we can introduce a desugaring from $\st_1 \rightharpoonup \st_2$ to $\sty{\rightharpoonup}{(\st_1, \st_2)}$.
+We can introduce a desugaring from $\st_1 \rightharpoonup \st_2$ to $\sty{\rightharpoonup}{(\st_1, \st_2)}$.
 %Here, we will benefit by treating it uniformly.
 
-All user-defined tycons $\tc$ must be defined in the \emph{tycon context}, $\Phi$, which is simply a list of tycon definitions, $\tcdef{\tc}{\psi}{\theta}$, where $\theta$ is the \emph{tycon structure} and $\psi$ is the \emph{tycon signature}.
-We will return to tycon structures below.
-Tycon signatures have the form $\tcsig{\ktyidx}{\chi}$, where $\ktyidx$ is the tycon's index kind and $\chi$ is the \emph{opcon signature}, which we   discuss in Sec.
-\ref{external-terms}.
-The first premise of \rulename{k-ty} extracts the index kind and the second checks the type index against it.
-
-The rule (k-otherty) governs types constructed by an ``other'' tycon.
-These will serve only as technical devices to stand in for types other than those in a given tycon context in Sec.
-\ref{metatheory}.
-The index of such a type must pair a natural number with a static value of kind $\kity$, discussed in Sec.
-\ref{sec:type-translations}.
-
-\paragraph{Examples} Two examples of tycon contexts (each defining only one tycon)  and their signatures are shown in Figure \ref{fig:example-tycons}.
-\footnote{In examples, we omit leading $\emptyset$, used as the base case for finite mappings, and $\cdot$, used as the base case for finite sequences, for concision.}
-
-Our first example, $\tcvar{rstr}$, has index kind $\krx$, which we assume classifies static regular expression patterns (defined as an inductive sum kind in the usual way).
-Types constructed by $\tcvar{rstr}$ will classify terms that behave as \emph{regular strings}, statically obeying the invariant that they are in the regular language specified by the type index \cite{sanitation-psp14}.
-For example, $\stx{title} := \sty{\tcvar{rstr}}{\concrx{.+}}$ will classify non-empty strings and $\stx{conf} := \sty{\tcvar{rstr}}{\concrx{[A-Z]+ \digit\digit\digit\digit}}$ will classify conference names.
-The type indices are  here written using standard concrete syntax for concision; recent work has specified how to define type-specific (or here, kind-specific) syntax like this composably in libraries \cite{TSLs}.
+\paragraph{Extension Types} For types constructed by an extension tycon, written in small caps, $\tc$, the rule \rulename{k-tc} extracts the index kind of $\tc$ from the tycon context, $\Phi$, which is simply a list of tycon definitions, $\tcdef{\tc}{\psi}{\theta}$, where $\theta$ is the \emph{tycon structure} and $\psi$ is the \emph{tycon signature}.
+Tycon signatures have the form $\tcsig{\ktyidx}{\chi}$, where $\ktyidx$ is the tycon's index kind.
+We cover the remaining components of tycon definitions and signatures as we continue.
+Two examples of tycon contexts (each defining only one tycon)\footnote{In examples, we omit leading $\emptyset$, used as the base case for finite mappings, and $\cdot$, used as the base case for finite sequences, for concision.} and their signatures are shown in Figure \ref{fig:example-tycons}.
+
+Our first example is $\tcvar{rstr}$.
+Its signature specifies that it has index kind $\krx$, which classifies static regular expression patterns (defined as an inductive sum kind in the usual way).
+Types constructed by $\tcvar{rstr}$ will classify terms that behave as \emph{regular strings}, i.e.
+they are statically known to be in the regular language specified by the type index \cite{sanitation-psp14}.
+For example, $\stx{title} := \sty{\tcvar{rstr}}{\concrx{.+}}$ will classify non-empty strings and $\stx{conf} := \sty{\tcvar{rstr}}{\concrx{[A-Z]+ \digit\digit\digit\digit}}$, shown in Sec.
+\ref{intro}, will classify conference names.
+The type indices are  here written using standard concrete syntax for concision; recent work has specified how to define type-specific (or here, kind-specific) syntax like this composably \cite{TSLs}.
 %We define the tycon context containing only the definition of $\tcvar{rstr}$, $\Phi_\text{rstr} := \tcdef{\tcvar{rstr}}{\tcsig{\krx}{\chi_\text{rstr}}}{\theta_\text{rstr}}$.
@@
 
-\subsection{Static Type Case Analysis}
+\paragraph{Other Types}
+The rule (k-otherty) governs types constructed by $\keyw{other}[m; \kappa]$.
+These will serve only as technical devices to stand in for types other than those in a given tycon context in Sec.
+\ref{metatheory}.
+The index must pair a term of kind $\kappa$  with a static term of kind $\kity$, discussed in Sec.
+\ref{sec:type-translations}.
+%The natural number $m$ serves to create arbitrarily many of these tycons.
+
+\subsection{Type Case Analysis}
 
@@
 \end{figure}
+\begin{figure}[t]
+\vspace{-10px}
+\begin{mathpar}
+\small\inferrule[tcc-ext]{
+    \vdash \Phi\\
+    \tc \notin \text{dom}(\Phi)\\
+    \keq{\emptyset}{\ktyidx}\\
+    \sofkz{\emptyset}{\emptyset}{\Phi}{\stx{schema}}{\ktyidx \rightarrow \kity}\\\\
+    \vdash_{\Phi,  \tcdef{\tc}{\tcsig{\ktyidx}{\chi}}{\tcstruct{\stx{schema}}{\omega}}} \omega \sim \tcsig{\ktyidx}{\chi}
+}{
+    \vdash \Phi, \tcdef{\tc}{\tcsig{\ktyidx}{\chi}}{\tcstruct{\stx{schema}}{\omega}}
+}\vspace{-10px}
+\end{mathpar}
+\caption{Tycon context well-definedness.}
+\label{fig:tycon-context-well-definedness}
+\vspace{-10px}
+\end{figure}
 
@@
 For example, the kinding rule for when $c$ is user-defined is below.
-\vspace{-5px}
 \begin{mathpar}
 \small
-\inferrule[k-tycase]{
+\inferrule[k-tccase]{
     \sofkX{\st}{\kty}\\
@@
 % \end{mathpar}
-The rule for $c{=}{\rightharpoonup}$ is analagous, but no rule for $c{=}\keyw{other}[m]$ is defined (such types always take the default branch).
-
-The dynamics (see supplement) are straightforwardly consistent with the intuition above.% We will see an example of its use in Sec. \ref{sec:targops}.
+The rule for $c{=}{\rightharpoonup}$ is analagous, but, importantly, no rule for $c{=}\keyw{other}[m; \kappa]$ is defined.
+%The dynamics (see supplement) are straightforwardly consistent with these intuitions.% We will see an example of its use in Sec.
+\ref{sec:targops}.
 
@@
 The tycon context well-definedness judgement, $\vdash \Phi$, requires that all tycon names are unique and performs three additional checks,  described below (we omit (tcc-emp)).
-\begin{mathpar}
-\small
-\inferrule[tcc-ext]{
-    \vdash \Phi\\
-    \tc \notin \text{dom}(\Phi)\\
-    \keq{\emptyset}{\ktyidx}\\
-    \sofkz{\emptyset}{\emptyset}{\Phi}{\stx{schema}}{\ktyidx \rightarrow \kity}\\\\
-    \vdash_{\Phi,  \tcdef{\tc}{\tcsig{\ktyidx}{\chi}}{\tcstruct{\stx{schema}}{\omega}}} \omega \sim \tcsig{\ktyidx}{\chi}
-}{
-    \vdash \Phi, \tcdef{\tc}{\tcsig{\ktyidx}{\chi}}{\tcstruct{\stx{schema}}{\omega}}
-}
-\end{mathpar}
 \subsection{Type Equivalence}\label{sec:type-equivalence}
-The first check simplifies the handling of type equivalence: type index kinds must be \emph{equality kinds}, i.e.
-those for which semantic equivalence implies syntactic equality at normal form.
+The first check simplifies type equivalence: type index kinds must be \emph{equality kinds}, i.e.
+those for which semantically equivalent values are syntactically equal.
 We define these by the judgement $\keq{\kDelta}{\kappa}$ (see supplement).
-Equality kinds are similar to equality types as found in Standard ML \cite{Tofte:89:TheDefinitionOfStandardML}.
-The main implication of this choice is that type indices cannot contain static functions.
+Arrow kinds are not equality kinds, so type indices cannot contain static functions.
+Equality kinds are analagous to equality types as found in Standard ML \cite{Tofte:89:TheDefinitionOfStandardML}.
 
@@
 \begin{array}{l}
-\slam{\klist{\klbl \times \kty}}{\svar{tyidx}}{\skap{\kity}{\skap{\klbl\times\kty}{\svar{listrec}}}~\svar{tyidx}~\sqity{\iunit}~\\
+\slam{\klist{\klbl \times \kty}}{\svar{tyidx}}{\svar{listrec}~\svar{tyidx}~\sqity{\iunit}~\\
   \quad (\lambda \svar{h}{:}\klbl \times \kty.\lambda \svar{r}{:}\kity.\sqity{\srep{{\ssnd{\svar{h}}}}\times\qtuq{\svar{r}}}}
@@
 Applying this translation schema to the index of $\stx{paper}$, for example, produces  $\stx{paper/trans} := \sqity{\qity_\text{paper/trans}}$ where $\small\qity_\text{paper/trans} := \srep{\stx{title}}\times(\srep{\stx{conf}}\times \iunit)$.
-Note that our schema did not remove the trailing unit type for simplicity (and to again emphasize that this choice will have only local implications due to the \emph{translation independence} property we will enforce as the core of our modularity guarantee).%For simplicity, and to make the point that the choice of representation has only tycon-local implications, our translation schema does not optimize away
+Note that our schema did not remove the trailing unit type for simplicity (and to again emphasize that this choice is local).%For simplicity, and to make the point that the choice of representation has only tycon-local implications, our translation schema does not optimize away
 
@@
 The judgement $\small\tdeabs{\Phi}{c}{\qity}{\memD}{\ity\moutput}{\memD\moutput}$ relates a normalized translational internal type $\qity$ to an internal type $\tau$, called a \emph{selectively abstracted type translation} because references to translations of types constructed by a tycon other than the ``delegated tycon'', $c$, are replaced by a corresponding type variable, $\alpha$.
-The \emph{type translation store} $\memD ::= \emptyset ~|~ \memD, \st \leftrightarrow \ity/\alpha$ maintains this correspondence between types, their actual translations and the unique type variables which appear in their place, e.g.
+The \emph{type translation store} $\memD ::= \emptyset ~|~ \memD, \st \leftrightarrow \ity/\alpha$ maintains this correspondence between types, their actual translations and the distinct type variables which appear in their place, e.g.
 $\small\tdeabs{\Phi_\text{rstr}\Phi_\text{lprod}}{\tcvar{lprod}}{\qity_\text{paper/trans}}{\emptyset}{\tau_\text{paper/abs}}{\memD_\text{paper/abs}}$
@@
 
-Each type translation store induces a \emph{type substitution}, $\delta$, and a corresponding internal type formation context, $\Delta$, according to the judgement $\memD \leadsto \delta : \Delta$.
-Type substitutions simply define an $n$-ary substitution for type variables, $\delta ::= \emptyset ~|~ \delta, \tau/\alpha$.
+The judgement $\memD \leadsto \delta : \Delta$ constructs the $n$-ary \emph{type substitution}, $\delta ::= \emptyset ~|~ \delta, \tau/\alpha$, and corresponding internal type formation context, $\Delta$, implied by the type translation store $\memD$.
 For example, $\memD_\text{paper/abs} \leadsto \delta_\text{paper/abs} : \Delta_\text{paper/abs}$ where $\delta_\text{paper/abs} := \keyw{str}/\alpha_1, \keyw{str}/\alpha_2$ and $\Delta_\text{paper/abs} := \alpha_1, \alpha_2$.
@@
 
-The rules for the selective type translation abstraction judgement are straightforward.
-We recurse generically over sub-terms of $\qity$ until sub-terms of form $\srep{\st}$ are encountered (see supplement).
-The translation of partial function types is direct and is not held abstract, so that lambdas can be used as the sole binding construct in the EL:
+The rules for the selective type translation abstraction judgement recurse generically over shared forms in $\qity$.
+Only sub-terms of form $\srep{\st}$ are interesting.
+The translations of function types are not held abstract, so that lambdas can serve as the sole binding construct in the EL:
 \begin{mathpar}
@@
 \end{mathpar}
-The translation of a user-defined type constructed by the delegated tycon is determined by calling the translation schema and checking that the type translation it generates refers only to type variables in $\memD'$:
+The translation of an extension type constructed by the delegated tycon is determined by calling the translation schema and checking that the type translation it generates is closed except for type variables tracked by $\memD'$:
 \begin{mathpar}
 \small
-\inferrule[abs-tc-local]{
+\inferrule[abs-tc-delegated]{
     \tcdef{\tc}{\psi}{\tcstruct{\stx{schema}}{\omega}} \in \Phi\\
@@
 \end{mathpar}
-The translation of a user-defined type constructed by any tycon other than the delegated tycon is added to the store (the supplement has the simple rule for retriving it once there):
+The translation of an extension type constructed by any tycon other than the delegated tycon is held abstract via a fresh type variable added to the store (the supplement has the rule (abs-stored) for retrieving it once there):
 \begin{mathpar}
 \small
-\inferrule[abs-tc-foreign-new]{
+\inferrule[abs-tc-not-delegated-new]{
     c \neq \tc\\
@@
 \end{mathpar}
-The translation of an ``other'' type is given directly in its index (rule (abs-other-foreign-new) is in the supplement):
+The translation of an ``other'' type is given in its index (rule (abs-other-not-delegated-new) is in the supplement):
 \begin{mathpar}
 \small
-\inferrule[abs-other-local]{
-    \tdeabs{\Phi}{\keyw{other}[m]}{\qity}{\memD}{\tau}{\memD'}
-}{
-    \tdeabs{\Phi}{\keyw{other}[m]}{\srep{\sty{\keyw{other}[m]}{(\stx{nat}, \sqity{\qity}}}}{\memD}{\tau}{\memD'}
+\inferrule[abs-other-delegated]{
+    \tdeabs{\Phi}{\keyw{other}[m;\kappa]}{\qity}{\memD}{\tau}{\memD'}
+}{
+    \tdeabs{\Phi}{\keyw{other}[m;\kappa]}{\srep{\sty{\keyw{other}[m;\kappa]}{(\stx{nat}, \sqity{\qity}}}}{\memD}{\tau}{\memD'}
 }
@@
  %We have that external typing contexts obey the standard structural congruences: weakening, exchange and contraction.
-\section{Typing and Translation of External Terms}\label{external-terms}
+\section{External Terms}\label{external-terms}
 
@@
     \stx{def}(\stx{args}) \Downarrow_{\es; \Upsilon; \Phi} \sqitm{\qitm}\\\\
-    \edeabs{\keyw{other}[m]}{\es; \Upsilon;\Phi}{\qitm}{\emptyset}{\emptyset}{\itm_\text{abs}}{\memD}{\memG}\\\\
-    \tdeabs{\Phi}{\keyw{other}[m]}{\srep{\sty{\keyw{other}[m]}{\sttyidx}}}{\memD}{\tau_\text{abs}}{\memD'}\\\\
+    \edeabs{\keyw{other}[m;\kappa]}{\es; \Upsilon;\Phi}{\qitm}{\emptyset}{\emptyset}{\itm_\text{abs}}{\memD}{\memG}\\\\
+    \tdeabs{\Phi}{\keyw{other}[m;\kappa]}{\srep{\sty{\keyw{other}[m;\kappa]}{\sttyidx}}}{\memD}{\tau_\text{abs}}{\memD'}\\\\
     \memD' \leadsto \delta : \Delta_\text{abs}\\
@@
 }{
-    \eanaX{\eintro{\stx{def}}{\es}}{\sty{\keyw{other}[m]}{\sttyidx}}{[\delta][\gamma]\iota_\text{abs}}
+    \eanaX{\eintro{\stx{def}}{\es}}{\sty{\keyw{other}[m;\kappa]}{\sttyidx}}{[\delta][\gamma]\iota_\text{abs}}
 }
@@
 \inferrule[syn-targ-other]{
-\esynX{e_\text{targ}}{\sty{\keyw{other}[m]}{\sttyidx}}{\iota_\text{targ}}\\\\
+\esynX{e_\text{targ}}{\sty{\keyw{other}[m;\kappa]}{\sttyidx}}{\iota_\text{targ}}\\\\
   \sofkz{\emptyset}{\emptyset}{\Phi}{\stx{def}}{\kargs\rightarrow(\kty\times\kitm)}\\\\
@@
   %\snorm{\srep{\sty{\tc}{\sttyidx}}}{\memD}{\cdot}{{\rightharpoonup}, \tc;\emptyset;\Phi}{\sqity{\ity_\text{abs}}}{\memD'}{\cdot}\\\\
-  \edeabs{\keyw{other}[m]}{{(e_\text{targ}; \es);\Upsilon;\Phi}}{\qitm}{\emptyset}{\emptyset}{\iota_\text{abs}}{\memD}{\memG}\\\\
-  \tdeabs{\Phi}{\keyw{other}[m]}{\srep{\st}}{\memD}{\ity_\text{abs}}{\memD'}\\\\
+  \edeabs{\keyw{other}[m;\kappa]}{{(e_\text{targ}; \es);\Upsilon;\Phi}}{\qitm}{\emptyset}{\emptyset}{\iota_\text{abs}}{\memD}{\memG}\\\\
+  \tdeabs{\Phi}{\keyw{other}[m;\kappa]}{\srep{\st}}{\memD}{\ity_\text{abs}}{\memD'}\\\\
   \memD' \leadsto \delta : \Delta_\text{abs}\\
@@
 The judgement $\edeabs{c}{\argEnv}{\qitm}{\memD}{\memG}{\itm\moutput}{\memD\moutput}{\memG\moutput}$, appearing as the eighth premise of (ana-intro), relates a translational internal term $\qitm$ to an internal term $\itm$ called a \emph{selectively abstracted term translation}, because all references to the translation of an argument (having any type) are replaced with a corresponding variable and, as in Sec.
-\ref{sec:selective-type-translation-abstraction}, all references to the translation of a type constructed by a user-defined tycon other than the ``delegated tycon'' $c$ are replaced with a corresponding abstract type variable.
+\ref{sec:selective-type-translation-abstraction}, all references to the translation of a type constructed by an extension tycon other than the ``delegated tycon'' $c$ are replaced with a corresponding abstract type variable.
 The type translation store, $\memD$, discussed previously, and term translation store, $\memG$, track these correspondences.
@@
 \subsection{Generalized Targeted Operations} 
-All non-introductory opcons associated with user-defined tycons go through another generalized form, in this case for \emph{targeted operations}, $\etarg{\opname{op}}{\sttmidx}{e_\text{targ}}{\es}$, where $\opname{op}$ ranges over opcon names, ${\sttmidx}$ is the term index, $e_\text{targ}$ is the \emph{target argument} and $\es$ are the remaining arguments.
+All non-introductory opcons associated with extension tycons go through another generalized form, in this case for \emph{targeted operations}, $\etarg{\opname{op}}{\sttmidx}{e_\text{targ}}{\es}$, where $\opname{op}$ ranges over opcon names, ${\sttmidx}$ is the term index, $e_\text{targ}$ is the \emph{target argument} and $\es$ are the remaining arguments.
 
@@
 
-% (k-ty)
+% (k-tc)
 
@@
 
-% (k-tycase-parr)
-
-% (k-tycase)
+% (k-tccase-parr)
+
+% (k-tccase)
 
@@
 \begin{proof-sketch} The proof is not unusually difficult because we have fixed the tycon context $\Phi_\text{rstr}$, so we can essentially treat the calculus like a type-directed compiler for a calculus with only two tycons, $\rightharpoonup$ and $\tcvar{rstr}$, plus one ``other'' one.
-Such a calculus and compiler specification was given in \cite{sanitation-psp14}, so we must simply show that the opcon definitions in $\tcvar{rstr}$ adequately capture these specifications using standard techniques for the SL, a simply-typed functional language \cite{conf/pldi/Chlipala07}.
+Such a calculus and compiler specification was given in \cite{sanitation-psp14}, so we must simply show that the opcon definitions in $\tcvar{rstr}$ adequately satisfy these specification using standard techniques for the SL, a simply-typed functional language \cite{conf/pldi/Chlipala07}.
 The only ``twist'' is that the rule (syn-targ-other) can synthesize a regular string type.
@@
 \begin{theorem}[Conservativity] If $\vdash \Phi$ and $\tc \in \text{dom}(\Phi)$ and a tycon invariant for $\tc$ holds under $\Phi$: \begin{itemize}
-\item If $\eana{\emptyset}{\Phi}{e}{\sty{\tc}{\sttyidx}}{\iota}$ and $\vdash_\Phi {\sty{\tc}{\sttyidx}} \leadsto \tau$  then $P(\sttyidx, \iota)$.
+\item If $\eana{\Upsilon}{\Phi}{e}{\sty{\tc}{\sttyidx}}{\iota}$ and $\vdash_\Phi \Upsilon \leadsto \Gamma$ and $\vdash_\Phi {\sty{\tc}{\sttyidx}} \leadsto \tau$  then $P(\Gamma, \sttyidx, \iota)$.
 \end{itemize} then for all $\Phi' = \Phi, \tcdef{\tc'}{\psi'}{\theta'}$ such that $\vdash \Phi'$, the same tycon invariant holds under $\Phi'$: \begin{itemize}
-\item If $\eana{\emptyset}{\Phi'}{e}{\sty{\tc}{\sttyidx}}{\iota}$ and $\vdash_{\Phi'}{\sty{\tc}{\sttyidx}}\leadsto{\tau}$ then $P(\sttyidx, \iota)$.
+\item If $\eana{\Upsilon}{\Phi'}{e}{\sty{\tc}{\sttyidx}}{\iota}$ and $\vdash_{\Phi'} \Upsilon \leadsto \Gamma$ and $\vdash_{\Phi'}{\sty{\tc}{\sttyidx}}\leadsto{\tau}$ then $P(\Gamma, \sttyidx, \iota)$.
 \end{itemize}
-(if proposition $P(\sttyidx, \iota)$ is \emph{modular}, defined below)
+(if proposition $P(\Gamma, \sttyidx, \iota)$ is \emph{modular}, defined below)
 \end{theorem}
@@
 $\tc$, the new term has a type constructed by that tycon with the same type translation, and only a slightly different type index.
-In particular, the mapping's effect on static terms is to replace all types constructed by $\tc'$ with a unique type constructed by $\keyw{other}[m]$ (for some suitable $m$ not already used).
-If $P(\sttyidx, \iota)$ is preserved under this transformation then we can simply invoke the existing proof of the tycon invariant.
+In particular, the mapping's effect on static terms is to replace all types constructed by $\tc'$ with a type constructed by $\keyw{other}[m;\ktyidx']$ for some $m$ corresponding to $\tc'$ and the index kind of $\tc'$.
+If $P(\Gamma, \sttyidx, \iota)$ is preserved under this transformation then we can simply invoke the existing proof of the tycon invariant.
 We call such propositions \emph{modular}.
@@
 describe a language-integrated technique specifically for this purpose, using a technique with parallels to the one we are building on \cite{TSLs}.
-Literal parsing logic is directly associated with user-defined types, forming \emph{type-specific languages} (TSLs), and local type inference controls invocation of TSL parsing logic, guaranteeing that composition is unambiguous.
+Literal parsing logic is directly associated with extension types, forming \emph{type-specific languages} (TSLs), and local type inference controls invocation of TSL parsing logic, guaranteeing that composition is unambiguous.
 The mechanism guarantees hygiene and inner terms cannot be inspected, so  fully modular reasoning is possible.
@@
 %When T, the solution is less clear. 
-When new distinctions are needed within an existing type, but new operators are not necessary, one solution is to develop an overlying system of \emph{refinement types} \cite{Freeman91}.
+When new static distinctions are needed within an existing type, but new operators are not necessary, one solution is to develop an overlying system of \emph{refinement types} \cite{Freeman91}.
 For example, a refinement of  integers might distinguish negative integers.
@@
 Although our regular string types could perhaps be approximated with refinements, this does not permit the advanced operations, optimizations and representation independence results we enable.
-Such a system could be seen in some ways as a degenerate mode of use of our work: we further cover the cases when new  operators are needed. 
+Were it to be, the system could be seen in some ways as a degenerate mode of use of our work: we further cover the cases when new  operators are needed. 
 For example, the $\tcvar{lprod}$ opcons use labels and types, which exist only statically.
@@
 We sidestep the most serious consequences of the EP by leaving our abstract syntax entirely fixed, instead  delegating to tycons.
-Few tools require knowledge of all external tycons in a typed translation semantics.
-As discussed previously, our treatment of concrete syntax in both the EL and SL defers to recent work on \emph{type-specific languages}, which takes a similar bidirectional approach for syntactic elaborations \cite{TSLs}.
-Some language frameworks do address the EP, but  provide few clear principles from which to reason modularly about metatheoretic properties specific to typed languages, like type safety and tycon invariants.
-One important decision that makes this tractable in @$\lambda$ is to associate opcons with tycons, forming \emph{modular tycons}, rather than considering terms and types both as open datatypes \cite{conf/ppdp/LohH06}.
+Fewer tools require knowledge of all external tycons in a typed translation semantics.
+As discussed previously, our treatment of concrete syntax in both the EL and SL defers to recent work on \emph{type-specific languages}, which takes a similar bidirectional approach for composably introducing syntactic desugarings \cite{TSLs}.
+Some language frameworks do address the EP, e.g.
+by encoding terms and types as open datatypes \cite{conf/ppdp/LohH06}, but this makes it quite difficult to reason modularly, particularly about metatheoretic properties specific to typed languages, like type safety and tycon invariants.
+Our key insight is to instead associate term-level opcons with tycons, which then become the fundamental constituents of the semantics (consistent with Harper's informal notation \cite{pfpl}).
 
@@
 
-A strength of @$\lambda$ is that fragment providers need not provide the semantics with mechanized proofs to modularly reason about language extension.
+In contrast, with @$\lambda$ fragment providers need not provide the semantics with mechanized proofs to benefit from modular reasoning principles.
 Instead, under a fixed tycon context, the calculus can be reasoned about like a very small type-directed compiler \cite{tarditi+:til-OLD,conf/pldi/Chlipala07}.
-Errors in reasoning can only lead to failure at typechecking time, a form of \emph{translation validation} \cite{Pnueli-Siegel-Singerman98}.
+Errors in reasoning can only lead to failure at typechecking time, via a novel form of \emph{translation validation} \cite{Pnueli-Siegel-Singerman98}.
 Incorrect opcon definitions (relative to a specification, e.g.
@@
 Thus, modular tycons can reasonably be tested ``in the field'' without concern about the reliability of the system as a whole.
-The simplicity of writing extensions directly in a functional language also suggests educational applications (following TinkerType \cite{LevinPierce99}).
-To eliminate even these localized failure modes for ``reliability-critical'' tycons, we plan to introduce \emph{optional} proof mechanization into the SL (by basing it on a dependently typed language like Coq) in subsequent work.
+To eliminate even these localized failure modes for ``reliability-critical'' tycons, we plan to introduce \emph{optional} proof mechanization into the SL (by basing it on a dependently typed language like Coq).
 %We also hope to mechanize the metatheory of @$\lambda$ itself in future work.%Every combination of fragments is a new dialect, and must be reasoned about monolithically.
@@
 Indeed, proofs of tycon invariants can rely on existing parametricity theorems \cite{WadlerThms}.
-Our work is technically reminiscent of  work on  translating modules to System $\mathbf{F}_\omega$ \cite{conf/tldi/RossbergRD10}.
+Our work is reminiscent of  work on elaborating an ML-style module system into System $\mathbf{F}_\omega$ \cite{conf/tldi/RossbergRD10}.
 Unlike in module systems, type translations (analagous to the choice of representation for an abstract type) are statically \emph{computed} based on a type index, rather than statically \emph{declared}.
@@
 In contrast, modules and ADTs can only specify a fixed number of operations, and each must have function type.
-Note also that these are not competing mechanisms: we did not specify quantification over external types here for simplicity, but we conjecture that it is complementary and thus @$\lambda$ could serve as the core of a language with an ML-style module system.
-Another related direction is \emph{tycon functors}, which would abstract over tycons with the same signature to support a tunable cost semantics.
+Note however that these are not competing mechanisms: we did not specify quantification over external types here for simplicity, but we conjecture that it is complementary and thus @$\lambda$ could serve as the core of a language with an ML-style module system.
+Another related direction is \emph{tycon functors}, which would abstract over tycons with the same signature to support tunable cost semantics.
 % This permits us tocode targeting a separate IL, and permit operations whose types cannot simply be written as functions.
@@
 Another limitation is that opcons cannot directly invoke one another (e.g.
-a \opname{len} opcon on regular strings could not construct a natural number; see supplement for a workaround, however).
+a \opname{len} opcon on regular strings could not construct a natural number).
 We conjecture that these are not fundamental limitations and expect @$\lambda$ to serve as a minimal foundation for future efforts that increase  expressiveness while maintaining the strong guarantees, like type safety and conservativity, established  here.

### 2014-11-04 att-pldi15/att-pldi15.tex
Updated PLDI15

@@
 row projection) using a static language where types and translations are values.
-The semantics guarantees \emph{type safety} and, unlike other approaches, \emph{conservativity}: that all \emph{tycon-specific invariants} that can be established in a ``closed world'' will be conserved in the ``open world''.
+The semantics guarantees \emph{type safety} and, critically, \emph{conservativity}: that all \emph{tycon-specific invariants} that can be established in a ``closed world'' will necessarily be conserved in the ``open world''.
 Type system providers need not provide mechanized proofs.
-Instead, these follow from a form of translation validation that uses type abstraction to check that tycons maintain \emph{translation independence}, analagous to the representation independence maintained by ML-style module systems.
+Instead, these guarantees require only a form of translation validation that uses type abstraction to check that tycons maintain \emph{translation independence}, analagous to representation independence in ML-style module systems.
 \end{abstract}
@@
 Universality properties (e.g.
-``Turing completeness'') are often enough to guarantee that an embedding that preserves a desirable fragment's dynamic semantics can be constructed, but an isomorphic embedding must also preserve the static semantics and, if defined, performance bounds specified using a cost semantics.
+``Turing completeness'') are often enough to guarantee that an embedding that preserves a desirable fragment's dynamic semantics can be constructed, but an isomorphic embedding must also preserve the static semantics and, if defined, performance bounds specified by a cost semantics.
 This is not always possible.
-For example, in $\mathcal{L}\{{\rightharpoonup}\,{\forall}\,{\mu}\,{\keyw{1}}\,{\times}\,{+}\}$, it is impossible to introduce record types as a library because you need the row projection operators, in ML written $\opname{\#}\conclbl{lbl}$, one for each of the infinite set of row labels $\conclbl{lbl}$ ($\opname{\#}$ itself is an \emph{operator constructor}).
-%You can only use $\mathcal{L}\{{\rightharpoonup}\,{\forall}\,{\mu}\,{\keyw{1}}\,{\times}\,{+}\}$ as a \emph{target of compilation} for a language with record types (they can be implemented using nested binary products,  lists, or other types).
+For example, in $\mathcal{L}\{{\rightharpoonup}\,{\forall}\,{\mu}\,{\keyw{1}}\,{\times}\,{+}\}$, it is impossible to introduce record types as a library because you need row projection operators, in ML written $\opname{\#}\conclbl{lbl}$, one for each of the infinite set of row labels $\conclbl{lbl}$ ($\opname{\#}$ itself is thus an \emph{operator constructor}).
+Each time a fragment like must be introduced directly, rather than as a library, a new \emph{dialect} of the language arises.
+Within the ML lineage, dialects that go beyond ``core ML'' abound:%, showing that this is not an isolated example:
+%Reynolds, in a remark that recalls the ``Turing tarpit'' of Perlis \cite{Perl82a}, summarizes the issue \cite{Reynolds94anintroduction}: 
+%\begin{quote}
+%To say that any reasonable function can be expressed by some program is not to say that it can be expressed by the most reasonable program.
+It is clear that the language requires a novel programming style.
+Moreover, it is likely that certain important functions cannot be expressed by their most efficient algorithms.
+%\end{quote}
+%%You can only use $\mathcal{L}\{{\rightharpoonup}\,{\forall}\,{\mu}\,{\keyw{1}}\,{\times}\,{+}\}$ as a \emph{target of compilation} for a language with record types (they can be implemented using nested binary products,  lists, or other types).
 %(and attempting to simulate records via a hash table over strings plus a refinement type system to check that the lookup operation will not fail will not satisfy a cost semantics prescribing $\mathcal{O}(1)$ lookup; see Sec.
@@
 %Indeed, the proliferation of language dialects constructed for this reason might be taken as an evidence that the core language is not as ``general'', when considered comprehensively, as might be hoped.
-Each time a fragment must be introduced directly, rather than as a library, a new \emph{dialect} of the language arises.
-Within the ML lineage, dialects that go beyond ``core ML'' abound:%, showing that this is not an isolated example:
-%Reynolds, in a remark that recalls the ``Turing tarpit'' of Perlis \cite{Perl82a}, summarizes the issue \cite{Reynolds94anintroduction}: 
-%\begin{quote}
-%To say that any reasonable function can be expressed by some program is not to say that it can be expressed by the most reasonable program.
-It is clear that the language requires a novel programming style.
-Moreover, it is likely that certain important functions cannot be expressed by their most efficient algorithms.
-%\end{quote}
-%
 
@@
 Mechanisms are needed that make it possible to define and  reason in a similarly modular, localized manner about  direct extensions to the semantics of a language.
-For example, if a language is extended with \emph{regular string types} as described in \cite{sanitation-psp14}, all terms having a regular string type like $\sty{\tcvar{rstr}}{\concrx{[A-Z]+ \digit\digit\digit\digit}}$ should continue to behave as strings in the regular language indicated no matter which other extensions are in use by a client.
+For example, if a language is extended with \emph{regular string types} as described in \cite{sanitation-psp14}, all terms having a regular string type like $\sty{\tcvar{rstr}}{\concrx{.+}}$ should continue to behave as non-empty strings no matter which other extensions are in use.
 %Such a mechanism could ultimately be integrated directly into the language, blurring the distinction between fragments and libraries and decreasing the need for new dialects.% Importing fragments that introduce new semantics   would be as easy and reasonable as importing a new module is in ML today.

### 2014-11-06 att-pldi15/att-pldi15-supplement.tex
Near final main paper

@@
 \end{mathpar}
+
+\begin{mathpar}\small
+\inferrule[s-itm-anatrans-v]{
+    \sval{\st}{\es;\Upsilon;\Phi}\\
+    \keyw{nth}[n](\es)=e
+}{
+    \sval{\sqitm{\anatrans{n}{\st}}}{\es;\Upsilon;\Phi}
+}
+
+\inferrule[s-itm-syntrans-v]{
+    \keyw{nth}[n](\es)=e
+}{
+    \sval{\sqitm{\syntrans{n}}}{\es;\Upsilon;\Phi}
+}
+\end{mathpar}
+
 We assume that the definitions of the standard helper functions $\small\svar{nil} :: \kforall{\kalpha}{\klist{\kalpha}}$ and $\small\svar{rcons} :: \kforall{\kalpha}{\klist{\kalpha} \rightarrow \kalpha \rightarrow \klist{\kalpha}}$, which adds an item to the end of a list, have been substituted into these rules.

### 2014-11-06 att-pldi15/att-pldi15.tex
Near final main paper

@@
 
-\title{Modularly Composing Typed Language Fragments}
+\title{Modular Type Constructors}
+\subtitle{Conservatively Composing Typed Language Fragments}
 
@@
 %,  Sometimes this fails because certain language features do not ``play well'' together, but precisely characterizing which is elusive given these practices.
-
 %Recent work has started to address this problem, e.g.
@@
 
-In this paper, we begin from first principles by specifying a language, @$\lambda$, in a style that follows many full-scale languages: as a bidirectionally typed translation semantics.
-Only the $\rightarrow$ type constructor (tycon) is built in; all other external tycons (we show constrained strings and a variation on record types) are defined by extending a \emph{tycon context}.
-Each tycon defines the semantics of its associated term-level \emph{opcons} (e.g.
+In this paper, we take a foundational approach, specifying an extensible typed translation semantics, @$\lambda$.
+Only the $\rightarrow$ type constructor is built in; all others (we discuss constrained strings and variants of record types) are defined by extending a \emph{tycon context}.
+Each tycon defines associated term-level \emph{opcons} (e.g.
 row projection) using a static language where types and translations are values.
-The semantics guarantees \emph{type safety} and, critically, \emph{conservativity}: that all \emph{tycon-specific invariants} that can be established in a ``closed world'' will necessarily be conserved in the ``open world''.
-Type system providers need not provide mechanized proofs.
-Instead, these guarantees require only a form of translation validation that uses type abstraction to check that tycons maintain \emph{translation independence}, analagous to representation independence in ML-style module systems.
+The semantics come with strong metatheoretic guarantees, notably \emph{type safety} and \emph{conservativity}: that \emph{tycon-specific  invariants} established in any ``closed world'' are conserved in the ``open world''.
+Remarkably, extension providers do not need to  provide the semantics with mechanized specifications or proofs.
+Instead, these guarantees arise from a form of translation validation that, taking inspiration from the ML module system, uses type abstraction to check that the opcons associated with a tycon respect the \emph{translation independence} of all others.% We can thus call them \emph{modular tycons}.
 \end{abstract}
@@
 \section{Introduction}\label{intro}
-Typed programming languages are most often described as being composed from \emph{fragments}, each contributing to the language's concrete syntax, abstract syntax, static semantics and dynamic semantics.
+Typed programming languages are generally described as being composed from \emph{fragments}, each contributing to the language's concrete syntax, abstract syntax, static semantics and dynamic semantics.
 In his textbook, Harper organizes such fragments around type constructors, describing each in a different chapter \cite{pfpl}.
@@
 This is not always possible.
-For example, in $\mathcal{L}\{{\rightharpoonup}\,{\forall}\,{\mu}\,{\keyw{1}}\,{\times}\,{+}\}$, it is impossible to introduce record types as a library because you need row projection operators, in ML written $\opname{\#}\conclbl{lbl}$, one for each of the infinite set of row labels $\conclbl{lbl}$ ($\opname{\#}$ itself is thus an \emph{operator constructor}).
-Each time a fragment like must be introduced directly, rather than as a library, a new \emph{dialect} of the language arises.
-Within the ML lineage, dialects that go beyond ``core ML'' abound:%, showing that this is not an isolated example:
+For example, in $\mathcal{L}\{{\rightharpoonup}\,{\forall}\,{\mu}\,{\keyw{1}}\,{\times}\,{+}\}$, it is impossible to introduce record types as a library because this requires introducing row projection operators,  written $\opname{\#}\conclbl{lbl}$ in ML, one for each of the infinite set of row labels $\conclbl{lbl}$ ($\opname{\#}$  is thus an \emph{operator constructor}).
+Each time such a situation occurs,  a new \emph{dialect} is needed.
+The ML lineage, like others, faces a proliferation of dialects:%, showing that this is not an isolated example:
 %Reynolds, in a remark that recalls the ``Turing tarpit'' of Perlis \cite{Perl82a}, summarizes the issue \cite{Reynolds94anintroduction}: 
@@
 \begin{enumerate}
-\vspace{-7px}
+\vspace{-5px}
 %\compresslist
@@
 labeled tuples, 
-records (identified up to reordering), 
-records with width and depth subtyping \cite{Cardelli:1984:SMI:1096.1098}, 
-records with update operators\footnote{The Haskell wiki notes that ''No, extensible records are not implemented in GHC.
+records (identified up to reordering), and 
+records with width and depth subtyping \cite{Cardelli:1984:SMI:1096.1098}, functional update operators\footnote{The Haskell wiki notes that ''No, extensible records are not implemented in GHC.
 The problem is that the record design space is large, and seems to lack local optima.
 [...] As a result, nothing much happens.'' \cite{GHCFAQ}} \cite{ocaml-manual},
-records with mutable fields \cite{ocaml-manual}, 
+ mutable fields \cite{ocaml-manual}, 
 %field delegation \cite{atlang-gpce14} \todo{gpce submission} 
 and 
-records with ``methods'' (i.e. pure objects \cite{TSLs}). 
- Sum-like types are also exposed variously: 
-standard datatypes, 
+ ``methods'' (i.e. pure objects) \cite{TSLs}. 
+ Sum-like types are also exposed variously: standard datatypes, 
 open datatypes \cite{conf/ppdp/LohH06,journals/toplas/MillsteinBC04}, 
 polymorphic variants \cite{ocaml-manual} and 
-exception types \cite{harper1997programming}. Combinations of these manifest themselves as class-based object systems \cite{ocaml-manual}.
+exception types \cite{harper1997programming}. Combinations occur as class-based object systems \cite{ocaml-manual}.
 
@@
 \textbf{Specialized Fragments:} Fragments that track specialized static invariants are also frequently introduced in dialects, e.g.
-for distributed programming \cite{Murphy:2007:TDP:1793574.1793585}, reactive programming \cite{mandel2005reactiveml}, authenticated data structures \cite{Miller:2014:ADS:2535838.2535851}, databases \cite{Ohori:2011:MSM:2034773.2034815},  units of measure \cite{conf/cefp/Kennedy09} and regular string sanitation \cite{sanitation-psp14}, amongst many other examples.
+for distributed programming \cite{Murphy:2007:TDP:1793574.1793585}, reactive programming \cite{mandel2005reactiveml}, databases \cite{Ohori:2011:MSM:2034773.2034815},  units of measure \cite{conf/cefp/Kennedy09} and regular string sanitation \cite{sanitation-psp14}, amongst many others.
 %\verb|sprintf| in the OCaml dialect statically distinguishes format strings from strings.% All of these are implemented as dialects of existing languages, presumably because a strong encoding was not feasible.
@@
 These then continue to hold no matter which other modules are in use by a client \cite{harper1997programming}.
-Other languages provide similar forms of \emph{abstract data types} to localize reasoning \cite{liskov1974programming}.
+Other languages expose similar forms of \emph{abstract data types} that also localize reasoning \cite{liskov1974programming}.
 %Mechanisms that can help decrease the complexity of an embedding without violating abstraction barriers are thus valuable, and we will lead into our work in Sec.
@@
 
-Mechanisms are needed that make it possible to define and  reason in a similarly modular, localized manner about  direct extensions to the semantics of a language.
+The alarming proliferation of dialects above suggests that mechanisms  that make it possible to define and  reason in a similarly modular, localized manner about  direct extensions to the semantics of a language are needed.
 For example, if a language is extended with \emph{regular string types} as described in \cite{sanitation-psp14}, all terms having a regular string type like $\sty{\tcvar{rstr}}{\concrx{.+}}$ should continue to behave as non-empty strings no matter which other extensions are in use.
@@
 Each tycon, e.g.
-$\tcvar{lprod}$ defining labeled products or $\tcvar{rstr}$ defining regular strings, determines the semantics of its associated opcons (e.g.
-$\opname{\#}$ for row projection, or $\opname{conc}$ for concatenation) via \emph{static functions}, i.e.
+$\tcvar{lprod}$ defining labeled products or $\tcvar{rstr}$ defining regular strings, determines the semantics of its associated opcons, e.g.
+$\opname{\#}$ for row projection, or $\opname{conc}$ for concatenation, via \emph{static functions}, i.e.
 functions written in a \emph{static language} (SL), where types and translations are values.
-
-All translation invariants maintained by these opcons in any ``closed world'', e.g.
-the regular string invariant just mentioned, are guaranteed to be maintained in any further extended tycon context, i.e.
+In other words, the semantics are controlled by a form of static code generation.
+
+We call a tycon associated with opcon definitions a \emph{modular tycon} because all translation invariants maintained by the opcons associated with a tycon in any ``closed world'', e.g.
+the regular string invariant just mentioned under the tycon context defining only $\tcvar{rstr}$, are necessarily maintained in any further extended tycon context, i.e.
 in the ``open world'', due to a simple static check that maintains \emph{translation independence} between tycons using type abstraction in the IL, the same fundamental mechanism underlying representation independence in ML-style module systems.
-As in ML, mechanized specifications and proofs are not needed.%, which provides a form of type-level computation.
+As in ML, mechanized specifications and proofs are not needed: the modularity guarantee is for all invariants.%, which provides a form of type-level computation.
 %The authors demonstrate the expressive power of this technique primarily with an implementation and a number of examples of fragments as libraries, and their core calculus .% (unlike systems that treat the type system as a ``bag of rules'', where non-determinism can arise).
@@
 \caption{Syntax of {$\mathcal{L}\{{\rightharpoonup}\,{\forall}\,{\mu}\,{\iunit}\,{\times}\,{+}\}$}, our internal language (IL).
-Metavariable $x$ ranges over term variables and $\alpha$ and $t$ both range over type variables.}
+Metavariable $x$ ranges over term variables and $\alpha$ and $t$ (distinguished only for stylistic reasons) over type variables.}
 \label{syntax-IL}
@@
 \end{figure}
+\paragraph{Internal Language} 
+%Internal terms, $\iota$, together with internal types, $\tau$, form the \emph{typed internal language}. 
+Our approach requires a typed internal language supporting type abstraction (i.e.
+universal quantification over types) \cite{Reynolds94anintroduction}.
+We use {$\mathcal{L}\{{\rightharpoonup}\,{\forall}\,{\mu}\,{\iunit}\,{\times}\,{+}\}$}, the syntax for which is shown in Figure \ref{syntax-IL}, as representative of a typical intermediate language for a typed language.
+%Note that the IL has a fixed semantics; changes to the IL induce different dialects of @$\lambda$.
+
+We assume the statics of the IL are specified in the standard way by judgements for  type formation {$\iDelta \vdash \tau$}, typing context formation { $\iDelta \vdash \iGamma$} and type assignment {$\iDelta~\iGamma \vdash \iota : \tau\moutput$}.
+%The typing and type formation contexts obey standard structural properties (i.e. weakening, exchange and contraction). 
+The internal dynamics are specified as a structural operational semantics with a stepping judgement {\small $\iota \mapsto \iota\moutput$} and a value judgement {$\iota~\mathtt{val}$}.
+Both the static and dynamic semantics of the IL can be found in any standard textbook covering typed lambda calculi (we directly follow \cite{pfpl}), so we assume familiarity and omit the details.
+
 \paragraph{External Language}
@@
 These are pronounced ``$e$ (synthesizes / analyzes against) type $\sigma$ and has  translation $\iota$ under typing context $\Upsilon$ and tycon context $\Phi$''.
-Note that our specifications in this paper are intended to be algorithmic: we indicate ``outputs'' when introducing judgement forms by \emph{mode annotations}, $\moutput$.
-Note that the type is an ``output'' only for the synthetic judgement.
-
-This separation of the EL from the IL is commonly used for full-scale language specifications, e.g.
-the Harper-Stone semantics for Standard ML \cite{Harper00atype-theoretic} or the semantics of Wyvern \cite{TSLs}.
-The internal language is purposely kept small, e.g.
+Our specifications are intended to be algorithmic: we indicate ``outputs'' when introducing judgement forms by \emph{mode annotations}, $\moutput$.
+Note that the type is an ``output'' only for the synthetic judgement.%, i.e.
+it can only be derived when the term itself has enough information in it to determine its type.
+
+We choose bidirectional typechecking, also sometimes called \emph{local type inference} \cite{Pierce:2000:LTI:345099.345100}, for two main reasons.
+The first is once again to justify the practicality of our approach: local type inference is increasingly being used in modern languages (e.g.
+Scala \cite{OdeZenZen01}) because it eliminates the need for type annotations in many situations while being simpler and decidable in more situations than whole-function type inference and providing what are widely perceived to be higher quality error messages \cite{journals/jfp/JonesVWS07}.
+Secondly, it will  give us a clean way to reuse the generalized intro form $\eintro{\st}{\es}$ and its desugarings at many types \cite{TSLs}.
+For example, regular string types can use standard string literal syntax and variants of record types can share   forms like $\{\conclbl{lbl1}=e_1, \conclbl{lbl2}=e_2\}$.
+
+Elaboration semantics distinguishing the EL from the IL have previously been found suitable for full-scale language specifications, e.g.
+the Harper-Stone semantics for Standard ML \cite{Harper00atype-theoretic}.
+The IL is purposely kept small, e.g.
 defining only binary products, to simplify metatheoretic reasoning and compilation.
@@
 In @$\lambda$, the EL builds in only function types.
-All other external constructs are defined in the \emph{tycon context}, $\Phi$, described starting in the Sec.
-\ref{types}.
-
-We choose bidirectional typechecking, also sometimes called \emph{local type inference} \cite{Pierce:2000:LTI:345099.345100}, for two main reasons.
-The first is once again to justify the practicality of our approach: local type inference is increasingly being used in modern languages (e.g.
-Scala \cite{OdeZenZen01}) because it eliminates the need for type annotations in many situations while being simpler and decidable in more situations than whole-function type inference and providing what are widely perceived to be higher quality error messages \cite{journals/jfp/JonesVWS07}.
-Secondly, it  gives us a clean way to reuse the generalized introductory form, $\eintro{\st}{\es}$, and its associated desugarings, at many types \cite{TSLs}.
-For example, regular string types can use standard string literal syntax.
-
-Unlike the Harper-Stone semantics, where external and internal terms were governed by a common type system, in @$\lambda$ each external type, $\st$, maps onto an internal type, $\tau$, called the \emph{type translation} of $\st$.
+All other external constructs will arise from the \emph{tycon context}, $\Phi$.
+
+This specification style is perhaps even more directly comparable to a specification for the initial stage of a type-directed compiler, e.g.
+the TIL compiler for Standard ML \cite{tarditi+:til-OLD}, here lifted ``one level up'' into the semantics of the language itself.
+The reason is that in the Harper-Stone elaboration semantics, external and internal terms were governed by a common type system, but in @$\lambda$ each type, $\st$, maps onto an internal type, $\tau$, called the \emph{type translation} of $\st$.
 This mapping is specified by the  type translation judgement, $\vdash_\Phi \st \leadsto \tau$, which will be described in Sec.
 \ref{sec:type-translations}.
-For example, regular string types will translate to internal type abbreviated $\keyw{str}$, and labeled product types will translate to nested binary product types, though we will emphasize that there are other valid choices.
-
-This specification style may thus also be compared to specifications for the first stage of a type-directed compiler, e.g.
-the TIL compiler for Standard ML \cite{tarditi+:til-OLD}, here lifted ``one level up'' into the semantics of the language itself.
-As we will see, type safety follows from a property analagous to a correctness condition that arises in typed compilers.
-Modular reasoning will be based on holding the type translation of $\st$ abstract ``outside'' the tycon.
-
-External typing contexts $\Upsilon$ map variables to types, so we also need the judgement $\vdash_\Phi \Upsilon \leadsto \Gamma$.
+For example, regular string types will translate to internal string types, abbreviated $\keyw{str}$, and labeled product types will translate to nested binary product types, though we will emphasize that there are other valid choices, and that the choice of type translation should have only local impact.
+%This will be maintained using a form of translation validation, which also has historically been studied in the literature on compilers \cite{Pnueli-Siegel-Singerman98}.
+
+External typing contexts, $\Upsilon$, map variables to types, so we also need the judgement $\vdash_\Phi \Upsilon \leadsto \Gamma$.
 
@@
 
-\paragraph{Internal Language} 
-%Internal terms, $\iota$, together with internal types, $\tau$, form the \emph{typed internal language}. 
-@$\lambda$ requires a typed internal language supporting type abstraction (i.e.
-universal quantification over types) \cite{Reynolds94anintroduction}.
-We use {$\mathcal{L}\{{\rightharpoonup}\,{\forall}\,{\mu}\,{\iunit}\,{\times}\,{+}\}$}, the syntax for which is shown in Figure \ref{syntax-IL}, as representative of a typical intermediate language for a typed functional language.
-%Note that the IL has a fixed semantics; changes to the IL induce different dialects of @$\lambda$.
-
-We assume the statics of the IL are specified in the standard way by judgements for  type formation {$\iDelta \vdash \tau$}, typing context formation { $\iDelta \vdash \iGamma$} and type assignment {$\iDelta~\iGamma \vdash \iota : \tau\moutput$}.
-%The typing and type formation contexts obey standard structural properties (i.e. weakening, exchange and contraction). 
-The internal dynamics are specified as a structural operational semantics with a stepping judgement {\small $\iota \mapsto \iota\moutput$} and a value judgement {$\iota~\mathtt{val}$}.
-The multi-step judgement $\iota \mapsto^{*} \iota\moutput$ is the reflexive, transitive closure of the stepping judgement and the evaluation judgement $\iota \Downarrow \iota'$ is defined iff $\iota \mapsto^{*} \iota'$ and $\iota'~\mathtt{val}$.
-Both the static and dynamic semantics of the IL can be found in any standard textbook covering typed lambda calculi (we directly follow \cite{pfpl}), so we assume familiarity and omit the details.
-
 % We also here define a syntax for simultaneous substitutions, of terms for variables, $\gamma$, and of types for polymorphic type variables, $\delta$, and assume standard judgements ensuring that these are valid with respect to a valid context, $\Delta \vdash \gamma : \Gamma$ and $\vdash \delta : \Delta$.
@@
 \hspace{-5px}\input{syntax-table-SL}
-\caption{Syntax of the static language (SL). Metavariable $\sx$ ranges over static term variables, $\kalpha$ and $\kvar$ over kind variables and $n$ over natural numbers.}\vspace{-5px}
+\caption{Syntax of the static language (SL). Metavariable $\sx$ ranges over static term variables, $\kalpha$ and $\kvar$ over kind variables and $n$ over natural numbers.}\vspace{-8px}
 \label{syntax-SL}
 \end{figure}
-The main novelty of @$\lambda$ is the \emph{static language}, which itself forms a typed lambda calculus where we use the word 
+The main novelty of @$\lambda$ is the \emph{static language} (SL), which itself forms a typed lambda calculus where  
 %External terms are classified by (external) \emph{types}. 
-\emph{kind}, $\kappa$, to refer to the ``types'' of \emph{static terms}, $\sigma$.
+\emph{kinds}, $\kappa$, serve as the ``types'' of \emph{static terms}, $\sigma$.
 The syntax of the SL is given in Figure \ref{syntax-SL}.
-The portion of the SL covered by the first row of kinds and static terms, some of which are elided for concision, forms an entirely standard total functional programming language consisting of total functions, universal quantification over kinds, inductive kinds, and products and sums  \cite{pfpl}.
-The reader can consider these as forming a total subset of ML \cite{harper1997programming} or a simply-typed subset of Coq \todo{citation}and we will assume standard conveniences from such languages (e.g.
-let bindings, pattern matching and inference of type parameters when applying polymorphic functions) in examples for concision.
-Only three new  kinds are needed for the SL to serve its role as the language used to   control typing and translation of the EL: $\kty$, classifying types (Sec.
-\ref{types}), $\kity$, classifying \emph{translational internal types} (Sec.
-\ref{sec:type-translations}) and $\kitm$, classifying \emph{translational internal terms} (Sec.
+The portion of the SL covered by the first row of kinds and static terms, some of which are elided for concision, forms an entirely standard total functional programming language consisting of total functions, polymorphic kinds, inductive kinds, and products and sums  \cite{pfpl}.
+The reader can consider these as forming a total subset of ML \cite{harper1997programming} or a simply-typed subset of Coq \todo{citation}and we will assume standard conveniences in examples (e.g.
+let bindings and inference of type parameters) for concision.
+Only three new  kinds are needed for the SL to serve its role as the language used to   control typing and translation: $\kty$, classifying types (Sec.
+\ref{types}), $\kity$, classifying \emph{quoted translational internal types} (used to compute type translations in Sec.
+\ref{sec:type-translations}) and $\kitm$, classifying \emph{quoted translational internal terms} (used to compute term translations in Sec.
 \ref{sec:introop}).
+The forms $\sana{n}{\st}$ and $\ssyn{n}$ will serve to link the dynamics of the SL with external type synthesis and analysis (Sec.
+\ref{external-terms}).
 
@@
 The multi-step judgement $\smanystep{\st}{\argEnv}{\st\moutput}$ is the reflexive, transitive closure of the stepping judgement.
-The normalization judgement $\seval{\st}{\argEnv}{\st'}$ is defined iff $\smanystep{\st}{\argEnv}{\st'}$ and $\sval{\st'}{\argEnv}$.
+The normalization judgement $\seval{\st}{\argEnv}{\st'}$ is derivable iff $\smanystep{\st}{\argEnv}{\st'}$ and $\sval{\st'}{\argEnv}$.
 %${\st}{\memD}{\memG}{\aCtx}{\st\moutput}{\memD\moutput}{\memG\moutput}$, where $\memD$, $\memG$ and $\aCtx$ are also technical devices that will be described later; they too can be ignored from the perspective of user code.
@@
 \small
-\inferrule[k-parr]{
+\inferrule[k-parr-ty]{
     \sofkX{\st}{\kprod{\kty}{\kty}}
@@
 
-\inferrule[k-tc]{
+\inferrule[k-ext-ty]{
     \tcdef{\tc}{\tcsig{\ktyidx}{\chi}}{\theta} \in \Phi\\
@@
 
-\inferrule[k-otherty]{
+\inferrule[k-other-ty]{
     \sofkX{\sttyidx}{\kappa \times \kity}
@@
 The introductory form for kind $\kty$ is $\sty{c}{\st}$, where $c$ is a \emph{tycon} and $\st$ is the \emph{type index}.
-The dynamics are simple: the index is eagerly normalized and errors propagate  (see supplement for the complete set of rules in this paper).
+The dynamics are simple: the index is eagerly normalized and errors propagate  (the supplement gives the full set of rules for this paper).
 
@@
 
-\paragraph{Arrow Types} The rule \rulename{k-parr} specifies that the type index of partial function types must be a pair of types.
+\paragraph{Function Types} The rule \rulename{k-parr-ty} specifies that the type index of partial function types must be a pair of types.
 We thus say that $\rightharpoonup$ has \emph{index kind} $\kty \times \kty$.
-We can introduce a desugaring from $\st_1 \rightharpoonup \st_2$ to $\sty{\rightharpoonup}{(\st_1, \st_2)}$.
+In practice, we would introduce a desugaring from $\st_1 \rightharpoonup \st_2$ to $\sty{\rightharpoonup}{(\st_1, \st_2)}$.
 %Here, we will benefit by treating it uniformly.
 
-\paragraph{Extension Types} For types constructed by an extension tycon, written in small caps, $\tc$, the rule \rulename{k-tc} extracts the index kind of $\tc$ from the tycon context, $\Phi$, which is simply a list of tycon definitions, $\tcdef{\tc}{\psi}{\theta}$, where $\theta$ is the \emph{tycon structure} and $\psi$ is the \emph{tycon signature}.
-Tycon signatures have the form $\tcsig{\ktyidx}{\chi}$, where $\ktyidx$ is the tycon's index kind.
-We cover the remaining components of tycon definitions and signatures as we continue.
-Two examples of tycon contexts (each defining only one tycon)\footnote{In examples, we omit leading $\emptyset$, used as the base case for finite mappings, and $\cdot$, used as the base case for finite sequences, for concision.} and their signatures are shown in Figure \ref{fig:example-tycons}.
-
-Our first example is $\tcvar{rstr}$.
-Its signature specifies that it has index kind $\krx$, which classifies static regular expression patterns (defined as an inductive sum kind in the usual way).
+\paragraph{Extension Types} For types constructed by an \emph{extension tycon}, written in small caps, $\tc$, the rule \rulename{k-ext-ty} extracts the index kind of $\tc$ from the tycon context, $\Phi$.
+
+For example,  $\tcvar{rstr}$ specifies the index kind $\krx$, which classifies static regular expression patterns (defined as an inductive sum kind in the usual way).
 Types constructed by $\tcvar{rstr}$ will classify terms that behave as \emph{regular strings}, i.e.
-they are statically known to be in the regular language specified by the type index \cite{sanitation-psp14}.
-For example, $\stx{title} := \sty{\tcvar{rstr}}{\concrx{.+}}$ will classify non-empty strings and $\stx{conf} := \sty{\tcvar{rstr}}{\concrx{[A-Z]+ \digit\digit\digit\digit}}$, shown in Sec.
-\ref{intro}, will classify conference names.
+strings statically known to be in the specified regular language \cite{sanitation-psp14}.
+For example, $\stx{title} := \sty{\tcvar{rstr}}{\concrx{.+}}$, which was shown in Sec.
+\ref{intro}, will classify non-empty strings and $\stx{conf} := \sty{\tcvar{rstr}}{\concrx{[A-Z]+ \digit\digit\digit\digit}}$ will classify strings having the format of a conference name.
 The type indices are  here written using standard concrete syntax for concision; recent work has specified how to define type-specific (or here, kind-specific) syntax like this composably \cite{TSLs}.
@@
 
-Our second example is the tycon $\tcvar{lprod}$, which will define a variant of labeled product type (labeled products are like record types, but maintain a row  ordering; record types are also definable in a manner discussed in the supplement, but maintaining an ordering simplifies our discussion).
+Our second example is $\tcvar{lprod}$, which defines a variant of labeled product type (labeled products are like record types, but maintain a row  ordering; records are definable in a manner discussed in Sec.
+\ref{sec:introop}, but an ordering simplifies our discussion).
 We choose the index kind of $\tcvar{lprod}$ to be $\klist{\klbl \times \kty}$, where list kinds are defined as inductive sums in the usual way, and $\klbl$ classifies static representations of row labels.
-In the tycon context containing both tycon definitions, $\Phi_\text{rstr}\Phi_\text{lprod}$, we can define a labeled product type classifying conference papers, $\stx{paper} := \sty{\tcvar{lprod}}{\{\conclbl{title} : \stx{title}, \conclbl{conf} : \stx{conf}\}}$.
-Note that $\istype{\stx{paper}}{\Phi_\text{rstr}\Phi_\text{lprod}}$ and we again use kind-specific syntax, in this case for $\klist{\klbl \times \kty}$.
+In a tycon context defining both tycon definitions, we can define a type classifying conference papers, $\stx{paper} := \sty{\tcvar{lprod}}{\{\conclbl{title} : \stx{title}, \conclbl{conf} : \stx{conf}\}}$.
+%Note that $\istype{\stx{paper}}{\Phi_\text{rstr}\Phi_\text{lprod}}$ and 
+We again use kind-specific syntax, in this case for $\klist{\klbl \times \kty}$.
+
+Two tycon contexts, each defining just one of the tycons just mentioned, are shown in Figure \ref{fig:example-tycons}.\footnote{In examples, we omit leading $\emptyset$, used as the base case for finite mappings, and $\cdot$, used as the base case for finite sequences, for concision.} We write the tycon context containing both  $\Phi_\text{rstr}\Phi_\text{lprod}$.
+The syntax of tycon contexts in Figure \ref{syntax-TC} specifies that they are simply lists of tycon definitions, $\tcdef{\tc}{\psi}{\theta}$.
+The tycon structure, $\theta$, contains a \emph{translation schema} (Sec.
+\ref{sec:type-translations}) and opcon definitions in an \emph{opcon structure}, $\omega$ (Sec.
+\ref{external-terms}).
+\emph{Tycon signatures}, $\psi$,  have the form $\tcsig{\ktyidx}{\chi}$, where $\ktyidx$ is the tycon's index kind and $\chi$ is the \emph{opcon signature}, defining index kinds for the opcons in the opcon structure.
+The rule (k-ext-ty) only needs the tycon index kind.
+
+The tycon context well-definedness judgement, written $\vdash \Phi$, requires that all tycon names are unique and performs three additional checks, shown in Figure \ref{fig:tycon-context-well-definedness} and described below (we omit the trivial base case, (tcc-emp)).
 
 \paragraph{Other Types}
-The rule (k-otherty) governs types constructed by $\keyw{other}[m; \kappa]$.
+The rule (k-otherty) governs types constructed by a tycon of the form $\keyw{other}[m; \kappa]$.
 These will serve only as technical devices to stand in for types other than those in a given tycon context in Sec.
@@
 \ref{sec:type-translations}.
+These can be thought of as a technical realization of the informal practice of including a ``catch-all'' or token base type (e.g.
+unit or nat) in a calculus.
 %The natural number $m$ serves to create arbitrarily many of these tycons.
@@
 \end{figure}
+\noindent
+Types here can be thought of as arising from a distinguished ``open datatype'' \cite{conf/ppdp/LohH06} defined by the tycon context.
+Consistent with this view, a type $\st$ can be case analyzed using $\stycase{c}{\st}{\sx}{\st_1}{\st_2}$.
+If the value of $\st$ is constructed by $c$, its type index is bound to $\sx$ and the branch $\st_1$ is taken.
+For totality, a default branch, $\st_2$, must also be provided.
+For example, the kinding rule for extension tycons is:
+\begin{mathpar}
+\small
+\inferrule[k-tccase]{
+    \sofkX{\st}{\kty}\\
+    \tcdef{\tc}{\tcsig{\ktyidx}{\chi}}{\theta} \in \Phi\\
+    \sofk{\kDelta}{\kGamma, \sx :: \ktyidx}{\Phi}{\st_1}{\kappa}\\
+    \sofkX{\st_2}{\kappa}
+}{
+    \sofkX{\stycase{\tc}{\st}{\sx}{\st_1}{\st_2}}{\kappa}
+}
+\end{mathpar}
+% \begin{mathpar}\small
+% \inferrule[s-tycase-fail-2]{ }{
+%     \sstep{\stycase{c}{\sotherty{m}{\tau}}{x}{\st_1}{\st_2}}{\argEnv}{\st_2}
+% }
+% \end{mathpar}
+The rule for $c{=}{\rightharpoonup}$ is analagous, but, importantly, no rule for case analyzing against $c{=}\keyw{other}[m; \kappa]$ is defined (these types thus always take the default branch).
+%The dynamics (see supplement) are straightforwardly consistent with these intuitions.
+We will see an example of its use in an opcon definition in Sec. \ref{sec:targops}.
+
 \begin{figure}[t]
@@
 \label{fig:tycon-context-well-definedness}
-\vspace{-10px}
+\vspace{-8px}
 \end{figure}
-
-Types in @$\lambda$ can be thought of as arising from a distinguished ``open datatype'' defined by the tycon context \cite{conf/ppdp/LohH06}.
-Consistent with this view, a type $\st$ can be case analyzed using $\stycase{c}{\st}{\sx}{\st_1}{\st_2}$.
-If the value of $\st$ is constructed by $c$, its type index is bound to $\sx$ and the branch $\st_1$ is taken.
-For totality, a default branch, $\st_2$, must also be provided.
-For example, the kinding rule for when $c$ is user-defined is below.
-\begin{mathpar}
-\small
-\inferrule[k-tccase]{
-    \sofkX{\st}{\kty}\\
-    \tcdef{\tc}{\tcsig{\ktyidx}{\chi}}{\theta} \in \Phi\\
-    \sofk{\kDelta}{\kGamma, \sx :: \ktyidx}{\Phi}{\st_1}{\kappa}\\
-    \sofkX{\st_2}{\kappa}
-}{
-    \sofkX{\stycase{\tc}{\st}{\sx}{\st_1}{\st_2}}{\kappa}
-}
-\end{mathpar}
-% \begin{mathpar}\small
-% \inferrule[s-tycase-fail-2]{ }{
-%     \sstep{\stycase{c}{\sotherty{m}{\tau}}{x}{\st_1}{\st_2}}{\argEnv}{\st_2}
-% }
-% \end{mathpar}
-The rule for $c{=}{\rightharpoonup}$ is analagous, but, importantly, no rule for $c{=}\keyw{other}[m; \kappa]$ is defined.
-%The dynamics (see supplement) are straightforwardly consistent with these intuitions.% We will see an example of its use in Sec.
-\ref{sec:targops}.
-
-\subsection{Tycon Context Well-Definedness}
-
+\noindent
 % \begin{figure}\hfill \fbox{$\vdash \Phi$}\vspace{-25px}
@@
 % \end{figure}
-The tycon context well-definedness judgement, $\vdash \Phi$, requires that all tycon names are unique and performs three additional checks,  described below (we omit (tcc-emp)).
+
 \subsection{Type Equivalence}\label{sec:type-equivalence}
-The first check simplifies type equivalence: type index kinds must be \emph{equality kinds}, i.e.
+\noindent The first check in (tcc-ext) simplifies type equivalence: type index kinds must be \emph{equality kinds}, i.e.
 those for which semantically equivalent values are syntactically equal.
@@
 Arrow kinds are not equality kinds, so type indices cannot contain static functions.
-Equality kinds are analagous to equality types as found in Standard ML \cite{Tofte:89:TheDefinitionOfStandardML}.
+Equality kinds are exactly analagous to equality types as found in Standard ML \cite{Tofte:89:TheDefinitionOfStandardML}.
 
 \subsection{Type Translations}\label{sec:type-translations}
-Recall that every type $\st$ must have a type translation, $\tau$.
-Each tycon in the tycon context {computes} translations for the types it constructs as a function of each type's index by specifying a \emph{translation schema} in the {tycon structure}, $\theta$.
+\noindent 
+Recall that every type $\st$ has a type translation, $\tau$.
+Extension tycons {compute} translations for the types they construct as a function of each type's index by specifying a \emph{translation schema} in the {tycon structure}, $\theta$.
 A tycon with index kind $\ktyidx$ must define a translation schema of kind $\ktyidx \rightarrow \kity$, checked by (tcc-ext).
 
-The kind $\kity$ has a single introductory form, $\sqity{\qity}$, where $\qity$ is a \emph{translational internal type}.
-Each form in the syntax for internal types, $\tau$,  corresponds to a form in the syntax of translational internal types, $\qity$.
-For example, our schema for $\tcvar{rstr}$ will simply choose to ignore the type index and translate regular strings to strings, of internal type abbreviated $\keyw{str}$.
-We abbreviate the corresponding translational internal type $\hat{\keyw{str}}$ and define $\stx{rstr/trans} := \slam{\krx}{\svar{tyidx}}{\sqity{\hat{\keyw{str}}}}$.
-The kinding and dynamics for shared forms proceed recursively, e.g.
+Terms of kind $\kity$ are introduced by a quotation form, $\sqity{\qity}$, where $\qity$ is a \emph{translational internal type}.
+Each form of internal type, $\tau$,  corresponds to a form of translational internal type, $\qity$.
+For example,  $\tcvar{rstr}$ ignores the type index and translate regular strings to strings, of internal type abbreviated $\keyw{str}$.
+Abbreviating the corresponding translational internal type $\hat{\keyw{str}}$, we define $\stx{rstr/trans} := \slam{\krx}{\svar{tyidx}}{\sqity{\hat{\keyw{str}}}}$.
+The kinding and dynamics for these shared forms is simply recursive (maintaining the quotation marker), e.g.
 \begin{mathpar}
@@
 
-The syntax for translational internal types additionally includes an ``unquote'' form,  $\qtuq{\st}$, so that they can be constructed compositionally, as well as a form, $\srep{\st}$, that allows one type's translation to refer to another: \begin{mathpar}\small
+The syntax for $\qity$ further includes an ``unquote'' form,  $\qtuq{\st}$, so that they can be constructed compositionally, as well as a form, $\srep{\st}$, that allows one type's translation to refer to another's: \begin{mathpar}\small
 \inferrule[k-ity-unquote]{
@@
 \end{mathpar}
-These forms are needed in the translation schema for $\tcvar{lprod}$, which generates nested binary product types (though we could also have used a list) by recursing over the type index and referring to the translations of the types therein.
-We assume the standard $\small\svar{listrec} :: \kforall{\kalpha_1}{\kforall{\kalpha_2}{\klist{\kalpha_1}\rightarrow\kalpha_2\rightarrow(\kalpha_1\rightarrow\kalpha_2\rightarrow\kalpha_2)\rightarrow\kalpha_2}}$ in defining $\stx{lprod/trans}:=$
+These forms are needed in the translation schema for $\tcvar{lprod}$, which generates nested binary product types (though we could also have used, e.g., an internal list) by folding over the type index and referring to the translations of the types therein.
+$\stx{lprod/trans}:=$
 \[\small
 \begin{array}{l}
-\slam{\klist{\klbl \times \kty}}{\svar{tyidx}}{\svar{listrec}~\svar{tyidx}~\sqity{\iunit}~\\
+\slam{\klist{\klbl \times \kty}}{\svar{tyidx}}{\svar{fold}~\svar{tyidx}~\sqity{\iunit}~\\
   \quad (\lambda \svar{h}{:}\klbl \times \kty.\lambda \svar{r}{:}\kity.\sqity{\srep{{\ssnd{\svar{h}}}}\times\qtuq{\svar{r}}}}
 \end{array}\]
+%We assume a standard $\small\svar{fold} :: \kforall{\kalpha_1}{\kforall{\kalpha_2}{\klist{\kalpha_1}\rightarrow\kalpha_2\rightarrow(\kalpha_1\rightarrow\kalpha_2\rightarrow\kalpha_2)\rightarrow\kalpha_2}}$ in defining this translation schema.
+
 Applying this translation schema to the index of $\stx{paper}$, for example, produces  $\stx{paper/trans} := \sqity{\qity_\text{paper/trans}}$ where $\small\qity_\text{paper/trans} := \srep{\stx{title}}\times(\srep{\stx{conf}}\times \iunit)$.
-Note that our schema did not remove the trailing unit type for simplicity (and to again emphasize that this choice is local).%For simplicity, and to make the point that the choice of representation has only tycon-local implications, our translation schema does not optimize away
+Note that our schema did not remove the trailing unit type for simplicity (and to again emphasize that this choice is local).
+It may be helpful to derive that $\stx{paper/trans} :: \kity$ by the rules above.%For simplicity, and to make the point that the choice of representation has only tycon-local implications, our translation schema does not optimize away
 
 \subsubsection{Selective Type Translation Abstraction}\label{sec:selective-type-translation-abstraction}
+\noindent
 References to type translations are maintained in values of kind $\kity$ like this to  allow us to selectively hold them abstract, which will be the key to translation independence.
 This can be thought of as analagous to the process in ML by which the true identity of an abstract type in a module is held abstract outside the module until after typechecking.
-The judgement $\small\tdeabs{\Phi}{c}{\qity}{\memD}{\ity\moutput}{\memD\moutput}$ relates a normalized translational internal type $\qity$ to an internal type $\tau$, called a \emph{selectively abstracted type translation} because references to translations of types constructed by a tycon other than the ``delegated tycon'', $c$, are replaced by a corresponding type variable, $\alpha$.
+The judgement $\small\tdeabs{\Phi}{c}{\qity}{\memD}{\ity\moutput}{\memD\moutput}$ relates a normalized translational internal type $\qity$ to an internal type $\tau$, called a \emph{selectively abstracted type translation} because references to translations of types constructed by a tycon other than the \emph{delegated tycon}, $c$, are replaced by a corresponding type variable, $\alpha$.
 The \emph{type translation store} $\memD ::= \emptyset ~|~ \memD, \st \leftrightarrow \ity/\alpha$ maintains this correspondence between types, their actual translations and the distinct type variables which appear in their place, e.g.
@@
 
-The judgement $\memD \leadsto \delta : \Delta$ constructs the $n$-ary \emph{type substitution}, $\delta ::= \emptyset ~|~ \delta, \tau/\alpha$, and corresponding internal type formation context, $\Delta$, implied by the type translation store $\memD$.
+%This process of selectively holding the translations of all types constructed by any tycon other than the ``delegated tycon'' abstract will be critical for conservativity to hold below.
+Here, the translation of the type is being held abstract, rather than the type itself, and type translations are computed from indices, rather than declared ``literally''.
+
+The judgement $\memD \leadsto \delta\moutput : \Delta\moutput$ constructs the $n$-ary \emph{type substitution}, $\delta ::= \emptyset ~|~ \delta, \tau/\alpha$, and corresponding internal type formation context, $\Delta$, implied by the type translation store $\memD$.
 For example, $\memD_\text{paper/abs} \leadsto \delta_\text{paper/abs} : \Delta_\text{paper/abs}$ where $\delta_\text{paper/abs} := \keyw{str}/\alpha_1, \keyw{str}/\alpha_2$ and $\Delta_\text{paper/abs} := \alpha_1, \alpha_2$.
@@
     \istype{\st}{\Phi}\\
-    \tdeabs{\Phi}{\tc}{\srep{\st}}{\emptyset}{\tau}{\memD}\\
+    \tdeabs{\Phi}{c}{\srep{\st}}{\emptyset}{\tau}{\memD}\\
     \memD \leadsto \delta : \Delta
@@
 
-%This process of selectively holding the translations of all types constructed by any tycon other than the ``delegated tycon'' abstract will be critical for conservativity to hold below.
-Here, the translation of the type is being held abstract, rather than the type itself, and type translations are computed from indices, rather than declared ``literally''.
-
-The rules for the selective type translation abstraction judgement recurse generically over shared forms in $\qity$.
-Only sub-terms of form $\srep{\st}$ are interesting.
-The translations of function types are not held abstract, so that lambdas can serve as the sole binding construct in the EL:
-\begin{mathpar}
-\small
-\inferrule[abs-parr]{
-    \tdeabs{\Phi}{c}{\srep{\st_1}}{\memD}{\ity_1}{\memD'}\\
-    \tdeabs{\Phi}{c}{\srep{\st_2}}{\memD'}{\ity_2}{\memD''}
-}{
-    \tdeabs{\Phi}{c}{\srep{\sty{\rightharpoonup}{(\st_1, \st_2)}}}{\memD}{\tau_1 \rightharpoonup \tau_2}{\memD''}
-}
-\end{mathpar}
-The translation of an extension type constructed by the delegated tycon is determined by calling the translation schema and checking that the type translation it generates is closed except for type variables tracked by $\memD'$:
+The rules for the selective type translation abstraction judgement recurse generically over shared forms in $\qity$. Only sub-terms of form $\srep{\st}$ are interesting. 
+The translation of an extension type  is determined by calling the translation schema and checking that the type translation it generates is closed except for type variables tracked by $\memD'$.
+If constructed by the delegated tycon, it is not held abstract:
 \begin{mathpar}
@@
     \tcdef{\tc}{\psi}{\tcstruct{\stx{schema}}{\omega}} \in \Phi\\
-    \sevalNA{\stx{schema}(\sttyidx)}{\sqity{\qity}}\\
+    \seval{\stx{schema}(\sttyidx)}{\cdot;\emptyset;\Phi}{\sqity{\qity}}\\\\
     \tdeabs{\Phi}{\tc}{\qity}{\memD}{\ity}{\memD'}\\
@@
 \end{mathpar}
-The translation of an extension type constructed by any tycon other than the delegated tycon is held abstract via a fresh type variable added to the store (the supplement has the rule (abs-stored) for retrieving it once there):
+Otherwise, it is held abstract via a fresh type variable added to the store (the supplement has the rule (abs-ty-stored) for retrieving it if already there):
 \begin{mathpar}
@@
     \tcdef{\tc}{\psi}{\tcstruct{\stx{schema}}{\omega}} \in \Phi\\\\
-    \sevalNA{\stx{schema}(\sttyidx)}{\sqity{\qity}}\\
+    \seval{\stx{schema}(\sttyidx)}{\cdot;\emptyset;\Phi}{\sqity{\qity}}\\
     \tdeabs{\Phi}{\tc}{\qity}{\memD}{\ity}{\memD'}\\\\
@@
 \end{mathpar}
-The translation of an ``other'' type is given in its index (rule (abs-other-not-delegated-new) is in the supplement):
+The translation of an ``other'' type is given in its index:
 \begin{mathpar}
@@
 }{
-    \tdeabs{\Phi}{\keyw{other}[m;\kappa]}{\srep{\sty{\keyw{other}[m;\kappa]}{(\stx{nat}, \sqity{\qity}}}}{\memD}{\tau}{\memD'}
+    \tdeabs{\Phi}{\keyw{other}[m;\kappa]}{\srep{\sty{\keyw{other}[m;\kappa]}{(\st, \sqity{\qity}}}}{\memD}{\tau}{\memD'}
 }
 \end{mathpar}
+Rule (abs-other-not-delegated-new) is in the supplement.
+
+The translations of function types are not held abstract, so that lambdas can be the sole binding construct in the EL:
+\begin{mathpar}
+\small
+\inferrule[abs-parr]{
+    \tdeabs{\Phi}{c}{\srep{\st_1}}{\memD}{\ity_1}{\memD'}\\
+    \tdeabs{\Phi}{c}{\srep{\st_2}}{\memD'}{\ity_2}{\memD''}
+}{
+    \tdeabs{\Phi}{c}{\srep{\sty{\rightharpoonup}{(\st_1, \st_2)}}}{\memD}{\tau_1 \rightharpoonup \tau_2}{\memD''}
+}
+\end{mathpar}
+
+% \paragraph{Summary}
+% In summary, types (e.g.
+$\stx{paper}$) are constructed by applying tycons to type indices.
+Extension tycons compute quoted type translations ($\stx{paper/trans}$), which then can determine many selectively abstracted type translations depending on the delegated tycon (e.g.
+$\tau_\text{paper/abs}$).
+From any such, the actual type translation is determined ($\tau_\text{paper}$).
 
@@
 \section{External Terms}\label{external-terms}
-
-Having established how types are constructed, and how they determine selectively abstracted and from there actual type translations, we can finally give the typing and translation rules for external terms, shown in Figure \ref{typing}.
+\noindent 
+Now that we have established how types are constructed and how selectively abstracted and from there actual type translations are computed by static functions, we are ready to give the semantics for external terms, shown in Figure \ref{typing}.
 
@@
   \stx{def}(\sttyidx)(\stmidx)(\stx{args}) \Downarrow_{\es;\Upsilon;\Phi} \sqitm{\qitm}\\\\
+  \trvalidate{\es;\Upsilon;\Phi}{\tc}{\qitm}{\srep{\sty{\tc}{\sttyidx}}}{\iota}
+  %\validate{1}{2{3}{4}{5}
   %\snorm{\stx{def}~\sttyidx~\stmidx~\stx{args}}{\emptyset}{\memG_0}{{\rightharpoonup}, \tc; \Upsilon; \Phi}{\sqitm{\itm_\text{abs}}}{\memD}{\memG}\\\\
   %\snorm{\srep{\sty{\tc}{\sttyidx}}}{\memD}{\cdot}{{\rightharpoonup}, \tc;\emptyset;\Phi}{\sqity{\ity_\text{abs}}}{\memD'}{\cdot}\\\\
-  \edeabs{\tc}{{\es;\Upsilon;\Phi}}{\qitm}{\emptyset}{\emptyset}{\iota_\text{abs}}{\memD}{\memG}\\
-  \memG \leadsto \gamma : \Gamma_\text{abs}\\\\
-  \tdeabs{\Phi}{\tc}{\srep{\sty{\tc}{\sttyidx}}}{\memD}{\ity_\text{abs}}{\memD'}\\
-  \memD' \leadsto \delta : \Delta_\text{abs}\\\\
-  \Delta_\text{abs}~\Gamma_\text{abs} \vdash \iota_\text{abs} : \tau_\text{abs}
-}{
-  \eanaX{\eintro{\stmidx}{\es}}{\sty{\tc}{\sttyidx}}{[{\delta}][{\gamma}]\iota}
+}{
+  \eanaX{\eintro{\stmidx}{\es}}{\sty{\tc}{\sttyidx}}{\iota}
 }
@@
   \stx{def}(\sttyidx)(\stmidx)(\stx{args}) \Downarrow_{(e_\text{targ}; \es);\Upsilon;\Phi} (\st, \sqitm{\qitm})\\\\
+  \trvalidate{(e_\text{targ}; \es);\Upsilon;\Phi}{\tc}{\qitm}{\srep{\st}}{\iota}
   %\snorm{\stx{def}~\sttyidx~\stmidx~\stx{args}}{\emptyset}{\memG_0}{{\rightharpoonup}, \tc; \Upsilon; \Phi}{\sqitm{\itm_\text{abs}}}{\memD}{\memG}\\\\
   %\snorm{\srep{\sty{\tc}{\sttyidx}}}{\memD}{\cdot}{{\rightharpoonup}, \tc;\emptyset;\Phi}{\sqity{\ity_\text{abs}}}{\memD'}{\cdot}\\\\
-  \edeabs{\tc}{{(e_\text{targ}; \es);\Upsilon;\Phi}}{\qitm}{\emptyset}{\emptyset}{\iota_\text{abs}}{\memD}{\memG}\\
-  \memG \leadsto \gamma : \Gamma_\text{abs}\\\\
-  \tdeabs{\Phi}{\tc}{\srep{\st}}{\memD}{\ity_\text{abs}}{\memD'}\\
-  \memD' \leadsto \delta : \Delta_\text{abs}\\\\
-  \Delta_\text{abs}~\Gamma_\text{abs} \vdash \iota_\text{abs} : \tau_\text{abs}
-}{
-    \esynX{\etarg{\opname{op}}{\sttmidx}{e_\text{targ}}{\es}}{\st}{[\delta][\gamma]\iota_\text{abs}}
+}{
+    \esynX{\etarg{\opname{op}}{\sttmidx}{e_\text{targ}}{\es}}{\st}{\iota}
 }
@@
     \stx{def}(\stx{args}) \Downarrow_{\es; \Upsilon; \Phi} \sqitm{\qitm}\\\\
-    \edeabs{\keyw{other}[m;\kappa]}{\es; \Upsilon;\Phi}{\qitm}{\emptyset}{\emptyset}{\itm_\text{abs}}{\memD}{\memG}\\\\
-    \tdeabs{\Phi}{\keyw{other}[m;\kappa]}{\srep{\sty{\keyw{other}[m;\kappa]}{\sttyidx}}}{\memD}{\tau_\text{abs}}{\memD'}\\\\
-    \memD' \leadsto \delta : \Delta_\text{abs}\\
-    \memG \leadsto \gamma : \Gamma_\text{abs}\\
-    \Delta_\text{abs}~\Gamma_\text{abs} \vdash \iota_\text{abs} : \tau_\text{abs}
-}{
-    \eanaX{\eintro{\stx{def}}{\es}}{\sty{\keyw{other}[m;\kappa]}{\sttyidx}}{[\delta][\gamma]\iota_\text{abs}}
+    \trvalidate{\es;\Upsilon;\Phi}{\keyw{other}[m;\kappa]}{\qitm}{\srep{\sty{\keyw{other}[m;\kappa]}{\sttyidx}}}{\iota}
+}{
+    \eanaX{\eintro{\stx{def}}{\es}}{\sty{\keyw{other}[m;\kappa]}{\sttyidx}}{\iota}
 }
@@
   \stx{def}(\stx{args}) \Downarrow_{(e_\text{targ}; \es);\Upsilon;\Phi} (\st, \sqitm{\qitm})\\\\
+  \trvalidate{(e_\text{targ}; \es);\Upsilon;\Phi}{\keyw{other}[m; \kappa]}{\qitm}{\srep{\st}}{\iota}
   %\snorm{\stx{def}~\sttyidx~\stmidx~\stx{args}}{\emptyset}{\memG_0}{{\rightharpoonup}, \tc; \Upsilon; \Phi}{\sqitm{\itm_\text{abs}}}{\memD}{\memG}\\\\
   %\snorm{\srep{\sty{\tc}{\sttyidx}}}{\memD}{\cdot}{{\rightharpoonup}, \tc;\emptyset;\Phi}{\sqity{\ity_\text{abs}}}{\memD'}{\cdot}\\\\
-  \edeabs{\keyw{other}[m;\kappa]}{{(e_\text{targ}; \es);\Upsilon;\Phi}}{\qitm}{\emptyset}{\emptyset}{\iota_\text{abs}}{\memD}{\memG}\\\\
-  \tdeabs{\Phi}{\keyw{other}[m;\kappa]}{\srep{\st}}{\memD}{\ity_\text{abs}}{\memD'}\\\\
-  \memD' \leadsto \delta : \Delta_\text{abs}\\
-  \memG \leadsto \gamma : \Gamma_\text{abs}\\
-  \Delta_\text{abs}~\Gamma_\text{abs} \vdash \iota_\text{abs} : \tau_\text{abs}
 }{
      \esynX{\etarg{\opname{op}}{\stx{def}}{e_\text{targ}}{\es}}{\st}{[\delta][\gamma]\iota_\text{abs}}
-}\vspace{-8px}
+}\vspace{-5px}
 \end{mathpar}
@@
 \end{figure*}
-Let us derive $\small\eana{\Upsilon_\text{ex}}{\Phi_\text{rstr}\Phi_\text{lprod}}{e_\text{ex}}{\stx{paper}}{\iota_\text{ex}}$ where $\small\Upsilon_\text{ex}  := title \Rightarrow \stx{title}$ and $\small e_\text{ex} := \{\conclbl{title}=title, \conclbl{conf}=\texttt{"EXMPL 2015"}\}$.
+
+Let us now derive $\small\eana{\Upsilon_\text{ex}}{\Phi_\text{rstr}\Phi_\text{lprod}}{e_\text{ex}}{\stx{paper}}{\iota_\text{ex}}$ where $\small\Upsilon_\text{ex}  := title \Rightarrow \stx{title}$ and $\small e_\text{ex} := \{\conclbl{title}=title, \conclbl{conf}=\texttt{"EXMPL 2015"}\}$.
 The  translation will be $\small\iota_\text{ex} := (title, (\texttt{"EXMPL 2015"}_\text{IL}, ()))$, where $\small\texttt{"EXMPL 2015"}_\text{IL}$ is an internal string (of internal type $\keyw{str}$).
@@
 In this example, this is $\tcvar{lprod}$.
-We will use this as the \emph{delegated tycon} in the final premises of the rule.
+We will use this as the {delegated tycon} in the final premises of the rule.
 
 The second premise extracts the \emph{intro term index kind}, $\klitidx$, from the \emph{opcon signature}, $\chi$, and the third premise checks the provided term index against this kind.
-This is simply the kind of term index expected by the tycon, e.g., $\tcvar{lprod}$ specifies $\klist{\klbl}$, so that it can use the labeled collection form, while $\tcvar{rstr}$ specifies an intro index kind of $\keyw{Str}$, so that it can use the string literal form.
+This is simply the kind of term index expected by the tycon, e.g.
+in Figure \ref{fig:example-tycons}, $\tcvar{lprod}$ specifies $\klist{\klbl}$, so that it can use the labeled collection form, while $\tcvar{rstr}$ specifies an intro index kind of $\keyw{Str}$, so that it can use the string literal form.
 %\[\small\begin{array}{lcl}
@@
 This is a static function that is applied, in the seventh premise, to determine whether the term is well-typed, raising an error if not or determining the translation of the term if so.
-The function has access to the type index, the term index and an interface to the list of arguments, and is kind is checked by the  judgement $\vdash_\Phi \omega \sim \psi$, which appeared as the final premise of the rule (tcc-ext) and is defined in Figure \ref{ocstruct}.
-For example, $\small\stx{rstr/intro}:=$
+The function has access to the type index, the term index and an interface to the list of arguments, and its kind is checked by the  judgement $\vdash_\Phi \omega \sim \psi$, which appeared as the final premise of the rule (tcc-ext) and is defined in Figure \ref{ocstruct}.
+For example, the intro opcon definition for $\tcvar{rstr}$ is $\small\stx{rstr/intro}:=$
 \[\small\begin{array}{l}
@@
 
-The only introductory form for kind $\kitm$ is $\sqitm{\qitm}$, where $\qitm$ is a \emph{translational internal term}.
-This form is analagous to the introductory form for kind $\kity$ described in Sec.
-\ref{sec:type-translations}, $\sqity{\qity}$.
-Each form in the syntax of $\iota$ has a corresponding form in the syntax for $\qitm$ and both the kinding rules and dynamics simply recurse through these in the same manner as shown in Sec.
+The introductory form for kind $\kitm$ is also a quotation form, $\sqitm{\qitm}$, where $\qitm$ is a \emph{translational internal term}.
+It is analagous to the form $\sqity{\qity}$ for $\kity$ in Sec.
+\ref{sec:type-translations}.
+Each form in the syntax of $\iota$ has a corresponding form in the syntax for $\qitm$ and the kinding rules and dynamics simply recurse through these in the same manner as in Sec.
 \ref{sec:type-translations}.
@@
 \quad \keyw{let}~\svar{inhabited}:\kunit=\svar{uniqmap}~\svar{tyidx}~\keyw{in}\\
-\quad \skap{\kitm}{\skap{\karg}{\skap{\klbl}{\skap{\klbl\times\kty}{\svar{listrec3}}}}}~\svar{tyidx}~\svar{tmidx}~\svar{args}~\sqitm{\itriv}\\
+\quad \svar{fold3}~\svar{tyidx}~\svar{tmidx}~\svar{args}~\sqitm{\itriv}\\
 \quad\quad \lambda \svar{rowtyidx}{:}\klbl\times\kty.\lambda \svar{rowtmidx}{:}\klbl.\lambda\svar{rowarg}{:}\karg.\lambda \svar{r}{:}\kitm.\\
@@
 The rest of this opcon definition folds over the three lists provided as input: the list mapping labels to types provided as the type index, the list of labels provided as the term index, and the list of argument interfaces.
-We assume a straightforward helper function, $\svar{listrec3}$, that raises an error if the three lists are not of the same length.
+We assume a straightforward helper function, $\svar{fold3}$, that raises an error if the three lists are not of the same length.
 The base case is the translational empty product.
-The recursive case first checks that the label provided in the term index matches the label provided in the type index, using a helper function $\svar{lbleq} :: \klbl \rightarrow \klbl \rightarrow \kunit$.
+The recursive case checks, for each row, that the label provided in the term index matches the label provided in the type index, using a helper function $\svar{lbleq} :: \klbl \rightarrow \klbl \rightarrow \kunit$.
 Then, we request type analysis of the corresponding argument, $\svar{rowarg}$, against the type in the type index, $\svar{rowty}$, by writing $\svar{ana}~\svar{rowarg}~\svar{rowty}$.
-Here, $\svar{ana}$ is a helper function defined below that triggers type analysis of the provided argument, producing a translational internal term  of the form $\sqitm{\anatrans{n}{\st}}$, where $n$ is the position of $\svar{rowarg}$ in $\svar{args}$ and $\st$ is the value of $\svar{rowty}$, upon success or  raising an error if not.
+Here, $\svar{ana}$ is a helper function defined below that triggers type analysis of the provided argument.
+If this succeeds, it evaluates to an translational internal term  of the form $\sqitm{\anatrans{n}{\st}}$, where $n$ is the position of $\svar{rowarg}$ in $\svar{args}$ and $\st$ is the value of $\svar{rowty}$.
+If analysis fails, it raises an error.
 The final line constructs a nested tuple based on this translation and the recursive result.
-Taken together, the translational internal term that will be generated for our example involving $e_\text{ex}$ above is $\small\qitm_\text{ex} := (\anatrans{0}{\stx{title}}, (\anatrans{1}{\stx{conf}}, ()))$.
-
-\paragraph{Argument Interface Lists} We define the kind of \emph{argument interfaces} as a simple product of functions, $\karg := (\kty \rightarrow \kitm) \times (\kunit \rightarrow \kty\times\kitm)$, one for analysis and the other for synthesis The helper functions $\svar{ana}$ and $\svar{syn}$ simply project the corresponding function out, $\svar{ana} := \slam{\karg}{\svar{arg}}{\keyw{fst}(\svar{arg})}$ and $\svar{syn}  :=  \slam{\karg}{\svar{arg}}{\keyw{snd}(\svar{arg})}$.
-
-The \emph{argument interface list} is a static list of length $n$ where the $i$th entry is $(\slam{\kty}{\svar{ty}}{\sana{i}{\svar{ty}}}, \slam{\kunit}{\_}{\ssyn{i}})$.
-It is generated by the judgement $\keyw{args}(n)=\stx{args}$, where $n$ is the length of the argument list, written $|\es|=n$.
+Taken together, the translational internal term that will be generated for our example involving $e_\text{ex}$ above is $\small\qitm_\text{ex} := (\anatrans{0}{\stx{title}}, (\anatrans{1}{\stx{conf}}, ()))$, i.e.
+it simply recalls that the two arguments were analyzed against $\stx{title}$ and  $\stx{conf}$, without yet inserting their translations directly.
+This is done by the \emph{translation validation judgement}, which occurs in the final premise. We describe argument interfaces and translation validation next.
+
+\subsection{Argument Interfaces} 
+\noindent
+The kind of \emph{argument interfaces} is $\karg := (\kty \rightarrow \kitm) \times (\kunit \rightarrow \kty\times\kitm)$, i.e.
+a  product of functions, one for analysis and the other for synthesis.
+The helpers $\svar{ana}$ and $\svar{syn}$ only project them out, e.g.
+$\svar{ana} := \slam{\karg}{\svar{arg}}{\keyw{fst}(\svar{arg})}$.
+To actually perform analysis or synthesis, we must provide a link between the dynamics of the static language and the typing rules.
+This is purpose of the static forms $\sana{n}{\st}$ and $\ssyn{n}$.
+We provide each opcon definition with an \emph{argument interface list}, which  a static list of length $n$ where the $i$th entry is $(\slam{\kty}{\svar{ty}}{\sana{i}{\svar{ty}}}, \slam{\kunit}{\_}{\ssyn{i}})$.
+It is generated by the judgement $\keyw{args}(n)=\stx{args}$, where $n$ is the length of the actual argument list, written $|\es|=n$.
 
@@
 
- Recall that the kinding judgement is indexed by $n$. This is an upper bound on the argument index of terms of the form $\sana{n}{\st}$ and $\ssyn{n}$. This is enforced in their kinding rules:
+Recall that the kinding judgement is indexed by $n$.
+This is an upper bound on the argument index of terms of the form $\sana{n}{\st}$ and $\ssyn{n}$, enforced in Figure \ref{fig:kinding-ana-syn}.
+Thus, if $\keyw{args}(n)= \stx{args}$ then $\sofkn{\emptyset}{\emptyset}{\Phi}{n}{\stx{args}}{\kargs}$.
+The rule (ocstruct-intro) ruled out writing either of these forms explicitly in an opcon definition by checking against the bound $n=0$.
+This is to prevent out-of-bounds errors: tycon providers can only access these forms via the argument interface list, which must have the correct length.
+
+\begin{figure}\vspace{-5px}
 \begin{mathpar}\small
@@
     \sofkX{\ssyn{n'}}{\kty\times\kitm}
-}
+}\vspace{-6px}
 \end{mathpar}
-Thus, if $\keyw{args}(n)= \stx{args}$ then $\sofkn{\emptyset}{\emptyset}{\Phi}{n}{\stx{args}}{\kargs}$.
-
-The rule (ocstruct-intro) ruled out writing either of these forms explicitly in an opcon definition by checking against the bound $n=0$.
-This is to prevent out-of-bounds errors: tycon providers do not write these forms directly, only accessing them via the argument interface list, which is guaranteed to have the correct length.
+\caption{Kinding for the SL-EL interface.}
+\label{fig:kinding-ana-syn}
+\vspace{-8px}
+\end{figure}
  %This is  safe because $n$ is only an upper bound, so it can be relaxed safely.
 
-These forms serve as the link between the dynamics of the static language and the statics of the external language.
 For $\sana{n}{\st}$, after normalizing $\st$, the argument environment, which contains the arguments themselves and the typing and tycon contexts, $\argEnv ::= \es; \Upsilon; \Phi$, is consulted to retrieve the $n$th argument and analyze it against $\st$.
-If this succeeds, the translational internal term $\sqitm{\anatrans{n}{\st}}$ is generated:
+If this succeeds, $\sqitm{\anatrans{n}{\st}}$ is generated:
 \begin{mathpar}\small
@@
 
-The kinding rules also prevent these translational internal term forms  from being well-kinded when $n = 0$. Like the form $\srep{\st}$, these  are retained in values of kind $\kitm$:
+The kinding rules also prevent these forms  from being well-kinded when $n = 0$. Like the form $\srep{\st}$, these  are retained in values of kind $\kitm$, as shown in $\qitm_\text{ex}$.
+
+\subsection{Translation Validation}
+\noindent
+The judgement $\trvalidate{\argEnv}{c}{\qitm}{\srep{\st}}{\itm}$, defined by a single rule in Figure \ref{fig:translation-validation} and appearing as the final premise of (ana-intro), can be pronounced ``translational internal term $\qitm$ generated by an opcon associated with $c$ under argument environment $\argEnv$ is a valid translation for an external term of type $\st$, so the actual translation $\iota$ is produced''.
+For example, $$\trvalidate{(title; \concstr{EXMPL 2015}); \Upsilon_\text{ex}; \Phi_\text{rstr}\Phi_\text{lprod}}{\tcvar{lprod}}{\qitm_\text{ex}}{\srep{\stx{paper}}}{\iota_\text{ex}}$$
+
+\noindent
+The purpose of this rule is to check that the generated translation will be well-typed \emph{no matter what the translations of types other than those constructed by $c$ are}.
+This \emph{translation independence} property will be the key to our conservativity theorem in Sec.
+\ref{metatheory}.
+
+The first premise generates the selectively abstracted type translation for $\st$ given that $c$ was the delegated tycon as described in Sec.
+\ref{sec:selective-type-translation-abstraction}.
+In our running example, this is $\tau_\text{abs/paper}$, i.e.
+$\alpha_0 \times (\alpha_1 \times \kunit)$.
+
+\begin{figure}
+\small\fbox{$\trvalidate{\argEnv}{c}{\qitm}{\srep{\st}}{\iota\moutput}$}\vspace{-5px}
 \begin{mathpar}\small
-\inferrule[s-itm-anatrans-v]{
-    \sval{\st}{\es;\Upsilon;\Phi}\\
-    \keyw{nth}[n](\es)=e
-}{
-    \sval{\sqitm{\anatrans{n}{\st}}}{\es;\Upsilon;\Phi}
-}
-
-\inferrule[s-itm-syntrans-v]{
-    \keyw{nth}[n](\es)=e
-}{
-    \sval{\sqitm{\syntrans{n}}}{\es;\Upsilon;\Phi}
-}
-\end{mathpar}
-
-\paragraph{Selectively Abstracted Term Translations} The reason for this is again because we will hold argument translations abstract by replacing them with variables.
-The judgement $\edeabs{c}{\argEnv}{\qitm}{\memD}{\memG}{\itm\moutput}{\memD\moutput}{\memG\moutput}$, appearing as the eighth premise of (ana-intro), relates a translational internal term $\qitm$ to an internal term $\itm$ called a \emph{selectively abstracted term translation}, because all references to the translation of an argument (having any type) are replaced with a corresponding variable and, as in Sec.
+\inferrule[validate-tr]{
+  \tdeabs{\Phi}{c}{\srep{\st}}{\emptyset}{\ity_\text{abs}}{\memD}\\
+  \edeabs{c}{{\es;\Upsilon;\Phi}}{\qitm}{\memD}{\emptyset}{\iota_\text{abs}}{\memD'}{\memG}\\
+  \memD' \leadsto \delta : \Delta_\text{abs}\\
+  \memG \leadsto \gamma : \Gamma_\text{abs}\\
+  \Delta_\text{abs}~\Gamma_\text{abs} \vdash \iota_\text{abs} : \tau_\text{abs}
+}{
+\trvalidate{\es;\Upsilon;\Phi}{c}{\qitm}{\srep{\st}}{[\delta][\gamma]\iota_\text{abs}}
+}\end{mathpar}\vspace{-9px}
+\caption{Translation Validation}
+\label{fig:translation-validation}\vspace{-8px}
+\end{figure}
+
+The judgement $\edeabs{c}{\argEnv}{\qitm}{\memD}{\memG}{\itm\moutput}{\memD\moutput}{\memG\moutput}$, appearing as the next premise, relates a translational internal term $\qitm$ to an internal term $\itm$ called a \emph{selectively abstracted term translation}, because all references to the translation of an argument (having any type) are replaced with a corresponding variable and, as in Sec.
 \ref{sec:selective-type-translation-abstraction}, all references to the translation of a type constructed by an extension tycon other than the ``delegated tycon'' $c$ are replaced with a corresponding abstract type variable.
@@
 Each entry can be read ``argument $n$ having type $\st$ and translation $\iota$ appears as variable $x$ with type $\tau$'', e.g.
-\[\small\edeabs{\tcvar{lprod}}{(title; \texttt{"EXMPL 2015"});\Upsilon;\Phi_\text{rstr}\Phi_\text{lprod}}{\qitm_\text{ex}}{\emptyset}{\emptyset}{\iota_\text{ex/abs}}{\memD_\text{ex/abs}}{\memG_\text{ex/abs}}\] where $\itm_\text{ex/abs} := (x_0, (x_1, ()))$ and $\memG_\text{ex/abs} := 0 : \stx{title} \leadsto title/x_0 : \alpha_0, 1 : \stx{conf} \leadsto \texttt{"EXMPL 2015"}_\text{IL}/x_1 : \alpha_1$ and  $\memD_\text{ex/abs}$ is $\memD_\text{paper/abs}$ from Sec.
+$\small\edeabs{\tcvar{lprod}}{(title; \texttt{"EXMPL 2015"});\Upsilon;\Phi_\text{rstr}\Phi_\text{lprod}}{\qitm_\text{ex}}{\memD_\text{paper/abs}}{\emptyset}{\iota_\text{ex/abs}}{\memD_\text{paper/abs}}{\memG_\text{ex/abs}}$ where $\itm_\text{ex/abs} := (x_0, (x_1, ()))$ and $\memG_\text{ex/abs} := 0 : \stx{title} \leadsto title/x_0 : \alpha_0, 1 : \stx{conf} \leadsto \texttt{"EXMPL 2015"}_\text{IL}/x_1 : \alpha_1$ and $\memD_\text{paper/abs}$ is from Sec.
 \ref{sec:selective-type-translation-abstraction}.
@@
 \ref{sec:selective-type-translation-abstraction}.
-The key rule for references to argument translations derived via analysis is below ($\syntrans{n}$ is analagous; the full rules are in the supplement).
-Note that we are rederiving the translation already determined in (s-ana-success) for simplicity (in practice, this might be cached):
+The key rule for references to argument translations derived via analysis is below ($\syntrans{n}$ is analagous; the full rules are in the supplement).% Note that we are rederiving the translation already determined in (s-ana-success) for simplicity (in practice, this might be cached):
 \begin{mathpar}\small
@@
 
-Like $\memD$, each $\memG$ induces an internal term substitution,  $\gamma ::= \emptyset ~|~ \gamma, \iota/x$, and corresponding internal typing context $\Gamma$ by the judgement $\memG \leadsto \gamma : \Gamma$, appearing as the ninth premise.
-In this case, $\small
-\gamma_\text{ex/abs} := title/x_0, \texttt{"EXMPL 2015"}_\text{IL}/x_1$ and $\small\Gamma_\text{ex/abs} := x_0 : \alpha_0, x_1 : \alpha_1$.
-
-The tenth premise of (ana-intro) determines a selectively abstracted type translation for the type provided for analysis, also using the tycon of the type provided for analysis as the delegated tycon and starting with the same store to ensure that equal types have equal type variables.
-In this case $\small\tau_\text{ex/abs} := \alpha_0 \times (\alpha_1 \times \kunit)$ (alpha-equivalent to $\tau_\text{paper/abs}$ in Sec.
-\ref{sec:selective-type-translation-abstraction}).
-The eleventh premise extracts a type substitution, $\delta_\text{ex/abs}$, and type formation context, $\Delta_\text{ex/abs}$, from $\memD_\text{ex/abs}$, again equivalent to $\delta_\text{conf/abs}$ and $\Delta_\text{conf/abs}$ from Sec.
+The third premise of (validate-tr) generates the type substitution and type formation contexts from the final type translation store,  $\memD' \leadsto \delta_\text{paper/abs} : \Delta_\text{paper/abs}$ as described in Sec.
 \ref{sec:selective-type-translation-abstraction}.
-
-Finally, the twelfth premise checks the abstracted term translation against the abstracted type translation.
-Here, we are checking $\Delta_\text{ex/abs}~\Gamma_\text{ex/abs} \vdash \iota_\text{ex/abs} : \tau_\text{ex/abs}$, i.e.:
+Like type translation stores, each term translation store induces an internal term substitution,  $\gamma ::= \emptyset ~|~ \gamma, \iota/x$, and corresponding internal typing context $\Gamma$ by the judgement $\memG \leadsto \gamma : \Gamma$, appearing as the next premise.
+Here, $\small
+\gamma_\text{ex/abs} := title/x_0, \texttt{"EXMPL 2015"}_\text{IL}/x_1$ and $\small\Gamma_\text{ex/abs} := x_0{:}\alpha_0, x_1{:}\alpha_1$.
+
+Finally, the fifth premise checks the abstracted term translation against the abstracted type translation according to the internal statics.
+Here, $\Delta_\text{paper/abs}~\Gamma_\text{ex/abs} \vdash \iota_\text{ex/abs} : \tau_\text{ex/abs}$, i.e.:
 \[\small(\alpha_0, \alpha_1)~(x_0 : \alpha_0, x_1 : \alpha_1) \vdash (x_0, (x_1, ())) : \alpha_0 \times (\alpha_1 \times \kunit)\]
-In other words, the translation of the labeled product $e_\text{ex}$ generated by $\tcvar{lprod}$ is checked with the references to term and type translations of regular strings replaced by variables and type variables, respectively.
+So, the translation of the labeled product $e_\text{ex}$ generated by $\tcvar{lprod}$ is checked with the references to term and type translations of regular strings replaced by variables and type variables, respectively.
 But because our  definition treated arguments parametrically, the check succeeds.% We  describe  an ill-behaved operator in Sec.
@@
 
-Applying the substitutions $\gamma_\text{ex/abs}$ and $\delta_\text{ex/abs}$ in the conclusion of the rule, we arrive at the actual term translation $\iota_\text{ex}$, defined previously.
+Applying the substitutions $\gamma_\text{ex/abs}$ and $\delta_\text{paper/abs}$ in the conclusion of the rule, we arrive at the actual term translation $\iota_\text{ex}$, defined previously.
 Note that $\iota_\text{ex}$ has type $\tau_\text{paper}$ under the translation of $\Upsilon_\text{ex}$, i.e.
@@
 
-The rule (ana-intro-other) is used to introduce terms of a type constructed by an ``other'' tycon.
-The term index, rather than the tycon context, directly specifies the static function that maps the arguments to a translation.
-In all other respects, it is analagous.
-It is used as a technical device in Sec.
-\ref{metatheory}.
-
-\subsection{Generalized Targeted Operations} 
-All non-introductory opcons associated with extension tycons go through another generalized form, in this case for \emph{targeted operations}, $\etarg{\opname{op}}{\sttmidx}{e_\text{targ}}{\es}$, where $\opname{op}$ ranges over opcon names, ${\sttmidx}$ is the term index, $e_\text{targ}$ is the \emph{target argument} and $\es$ are the remaining arguments.
-
-Concrete desugarings include $e_\text{targ}.\opname{op}\langle\sttmidx\rangle(\es)$ (and variants where the term index or arguments are omitted), projection syntax for use by record-like types, $e_\text{targ}\opname{\#}\conclbl{lbl}$, which desugars to $\etarg{\opname{prj}}{\conclbl{lbl}}{e_\text{targ}}{\cdot}$, and $e_\text{targ} \cdot e_\text{arg}$, which desugars to $\etarg{\opname{conc}}{\striv}{e_\text{targ}}{e_\text{arg}}$.
+\subsection{Generalized Targeted Operations} \label{sec:targops}
+All non-introductory operations go through another generalized form, in this case for \emph{targeted operations}, $\etarg{\opname{op}}{\sttmidx}{e_\text{targ}}{\es}$, where $\opname{op}$ ranges over opcon names, ${\sttmidx}$ is the term index, $e_\text{targ}$ is the \emph{target argument} and $\es$ are the remaining arguments.
+Concrete desugarings for this form include $e_\text{targ}.\opname{op}\langle\sttmidx\rangle(\es)$ (and variants where the term index or arguments are omitted), projection syntax for use by record-like types, $e_\text{targ}\opname{\#}\conclbl{lbl}$, which desugars to $\etarg{\opname{\#}}{\conclbl{lbl}}{e_\text{targ}}{\cdot}$, and $e_\text{targ} \cdot e_\text{arg}$, which desugars to $\etarg{\opname{conc}}{\striv}{e_\text{targ}}{e_\text{arg}}$.
 We show other desugarings, including case analysis, in the supplement.
@@
 The opcons associated with $\tcvar{rstr}$ are taken directly from Fulton et al.'s specification of regular string types \cite{sanitation-psp14}, with the exception of $\opname{case}$, which generalizes case analysis as defined there to arbitrary string patterns, which we discuss in the supplement.
-The opcons associated with $\tcvar{lprod}$ are also straightforward: $\opname{prj}$ projects out the row with the provided label, $\opname{conc}$ concatenates two labeled products (updating common rows with the value from the right argument), and $\opname{drop}$ creates a new labeled product from the target with some rows dropped.
+The opcons associated with $\tcvar{lprod}$ are also straightforward: $\opname{\#}$ projects out the row with the provided label and $\opname{conc}$ concatenates two labeled products (updating common rows with the value from the right argument).
 Note that both $\tcvar{rstr}$ and $\tcvar{lprod}$ can define concatenation without conflict.
@@
 The rule (ocstruct-targ) in Figure \ref{ocstruct} ensures that no tycon defines an opcon twice and that the opcon definitions are well-kinded.
-For example, $\stx{rstr/conc}=$
+For example, $\stx{rstr/conc} :=$
 \[\small
@@
             \quad \keyw{letpair}~(\svar{arg1}, \svar{arg2}) = \svar{arity2}~\svar{args}~\keyw{in}\\
-            \quad \keyw{letpair}~(\_, \svar{tr1}) = \svar{syn}~\svar{arg1}~\keyw{in}~ \keyw{letpair}~(\svar{ty2}, \svar{tr2}) = \svar{syn}~\svar{arg2}\\
+            \quad \keyw{letpair}~(\_, \svar{tr1}) = \svar{syn}~\svar{arg1}~\keyw{in}~  \\\quad\keyw{letpair}~(\svar{ty2}, \svar{tr2}) = \svar{syn}~\svar{arg2}\\
             \quad \stycase{\tcvar{rstr}}{\svar{ty2}}{\svar{tyidx2}}{\\
@@
 We case analyze the second against $\tcvar{rstr}$, extracting its index regular expression if so and raising an error if not.
-We then synthesize the resulting regular string type, using the helper function $\svar{rxconcat} :: \krx \rightarrow \krx \rightarrow \krx$ which generates the synthesized type index, and finally the translation, using an internal helper function $sconcat : \keyw{str} \rightarrow \keyw{str} \rightarrow \keyw{str}$, the translational term for which we assume has been substituted in directly.
-
-The remaining premises of (syn-targ) are analagous to the corresponding premises in (ana-intro), with the only difference being that we check the selectively abstracted  term translation against the selectively abstracted type translation of the synthesized type, but the delegated tycon is that of the type synthesized by the target argument.
+We then synthesize the resulting regular string type, using the helper function $\svar{rxconcat} :: \krx \rightarrow \krx \rightarrow \krx$ which generates the synthesized type index by concatenating the indices of the argument's types, and finally the translation, using an internal helper function $sconcat : \keyw{str} \rightarrow \keyw{str} \rightarrow \keyw{str}$, the translational term for which we assume has been substituted in directly.
+
+The last premise of (syn-targ) again performs translation validation as described above.
+The only difference relative to (ana-intro) is that that we check the term translation against the type translation of the synthesized type, but the delegated tycon is that of the type synthesized by the target argument.
+
+\subsection{Operations Over Other Types}
+The rule (ana-intro-other) is used to introduce terms of a type constructed by an ``other'' tycon.
+The term index, rather than the tycon context, directly specifies the static function that maps the arguments to a translation.
+In all other respects, it is analagous.
+It is used as a technical device in Sec.
+\ref{metatheory}.
 
 Like (ana-intro-other), rule (syn-targ-other) is used when the target synthesizes an ``other'' type.
-The mapping from the arguments to a type and translation is given directly in the term index (the op name is ignored).
+The mapping from the arguments to a type and translation is again given directly in the term index (the op name is ignored).
 
@@
 \begin{theorem}[Static Progress]\label{thm:static-progress}
-If $\sofkn{\emptyset}{\emptyset}{\Phi}{n}{\st}{\kappa}$ and $\vdash \Phi$ and $|\es|=n$ and $\vdash_\Phi \Upsilon \leadsto \Gamma$ then $\sval{\st}{\es; \Upsilon; \Phi}$ or $\serr{\st}{\es; \Upsilon; \Phi}$ or $\sstep{\st}{\es; \Upsilon; \Phi}{\st'}$.
+If $\sofkn{\emptyset}{\emptyset}{\Phi}{n}{\st}{\kappa}$ and $\vdash \Phi$ and $|\es|=n$ then $\sval{\st}{\es; \Upsilon; \Phi}$ or $\serr{\st}{\es; \Upsilon; \Phi}$ or $\sstep{\st}{\es; \Upsilon; \Phi}{\st'}$.
 \end{theorem}
@@
 If $\sofkn{\emptyset}{\emptyset}{\Phi}{n}{\st}{\kappa}$ and $\vdash \Phi$ and $|\es|=n$ and $\vdash_\Phi \Upsilon \leadsto \Gamma$ and $\sstep{\st}{\es; \Upsilon; \Phi}{\st'}$ then $\sofkn{\emptyset}{\emptyset}{\Phi}{n}{\st'}{\kappa}$.
-\end{theorem}
-The case in the proof of Theorem \ref{thm:static-preservation} for $\ssyn{n}$ requires that the following theorem be mutually defined.
+\end{theorem}\noindent
+The case in the proof of Theorem \ref{thm:static-preservation} for $\st=\ssyn{n}$ requires that the following theorem be mutually defined.
 %The mutual induction is well-founded because the total number of intro and targeted sub-terms being considered decreases.
@@
 Type safety for the IL \cite{pfpl} then implies that evaluation cannot go wrong.
-To prove this, we must prove a stronger theorem, \emph{type-preserving translation}, analagous to type-preserving compilation in the typed compilation literature \cite{tarditi+:til-OLD}:% produced as the translation of the external type.
+To prove this, we must in fact prove a stronger theorem: that a term's translation has its type's translation (the analagous notion is \emph{type-preserving compilation} in type-directed compilers \cite{tarditi+:til-OLD}):% produced as the translation of the external type.
 Note that we need only to state our top-level theorems for the analysis judgement  because of the subsumption rule.
@@
 \begin{theorem}[Type-Preserving Translation]
-If $\vdash \Phi$ and $\vdash_\Phi \Upsilon \leadsto \Gamma$ and $\vdash_\Phi \st \leadsto \tau$ and $\eanaX{e}{\st}{\iota}$ then $\emptyset~\Gamma \vdash \iota : \tau$.
+If $\vdash \Phi$ and $\vdash_\Phi \Upsilon \leadsto \Gamma$ and $\vdash_\Phi \st \leadsto \tau$ and $\eanaX{e}{\st}{\iota}$ then $\emptyset \vdash \Gamma$ and $\emptyset \vdash \tau$ and $\emptyset~\Gamma \vdash \iota : \tau$.
 \end{theorem}
@@
 \begin{theorem}[Conservativity] If $\vdash \Phi$ and $\tc \in \text{dom}(\Phi)$ and a tycon invariant for $\tc$ holds under $\Phi$: \begin{itemize}
-\item If $\eana{\Upsilon}{\Phi}{e}{\sty{\tc}{\sttyidx}}{\iota}$ and $\vdash_\Phi \Upsilon \leadsto \Gamma$ and $\vdash_\Phi {\sty{\tc}{\sttyidx}} \leadsto \tau$  then $P(\Gamma, \sttyidx, \iota)$.
+\item For all $\Upsilon, e, \sttyidx$, if $\eana{\Upsilon}{\Phi}{e}{\sty{\tc}{\sttyidx}}{\iota}$ and $\vdash_\Phi \Upsilon \leadsto \Gamma$ and $\vdash_\Phi {\sty{\tc}{\sttyidx}} \leadsto \tau$  then $P(\Gamma, \sttyidx, \iota)$.
 \end{itemize} then for all $\Phi' = \Phi, \tcdef{\tc'}{\psi'}{\theta'}$ such that $\vdash \Phi'$, the same tycon invariant holds under $\Phi'$: \begin{itemize}
-\item If $\eana{\Upsilon}{\Phi'}{e}{\sty{\tc}{\sttyidx}}{\iota}$ and $\vdash_{\Phi'} \Upsilon \leadsto \Gamma$ and $\vdash_{\Phi'}{\sty{\tc}{\sttyidx}}\leadsto{\tau}$ then $P(\Gamma, \sttyidx, \iota)$.
+\item For all $\Upsilon, e, \sttyidx$, if $\eana{\Upsilon}{\Phi'}{e}{\sty{\tc}{\sttyidx}}{\iota}$ and $\vdash_{\Phi'} \Upsilon \leadsto \Gamma$ and $\vdash_{\Phi'}{\sty{\tc}{\sttyidx}}\leadsto{\tau}$ then $P(\Gamma, \sttyidx, \iota)$.
 \end{itemize}
-(if proposition $P(\Gamma, \sttyidx, \iota)$ is \emph{modular}, defined below)
+(if proposition $P(\Gamma, \st, \iota)$ is \emph{modular}, defined below)
 \end{theorem}
@@
 In particular, the mapping's effect on static terms is to replace all types constructed by $\tc'$ with a type constructed by $\keyw{other}[m;\ktyidx']$ for some $m$ corresponding to $\tc'$ and the index kind of $\tc'$.
-If $P(\Gamma, \sttyidx, \iota)$ is preserved under this transformation then we can simply invoke the existing proof of the tycon invariant.
+If $P(\Gamma, \st, \iota)$ is preserved under this transformation then we can simply invoke the existing proof of the tycon invariant.
 We call such propositions \emph{modular}.
-Non-modular propositions are uninteresting because they have knowledge of tycons ``from the future''.
+Non-modular propositions are uninteresting because they distinguish tycons ``from the future''.
 
@@
 Although our regular string types could perhaps be approximated with refinements, this does not permit the advanced operations, optimizations and representation independence results we enable.
-Were it to be, the system could be seen in some ways as a degenerate mode of use of our work: we further cover the cases when new  operators are needed. 
-For example, the $\tcvar{lprod}$ opcons use labels and types, which exist only statically.
-Thus, labeled tuples cannot be seen as refinements of nested pairs because the row projection operators don't exist.
+Were it to be, the system could be seen in ways as a degenerate mode of use of our work: we further cover the cases when new  operators are needed. 
+For example, labeled tuples couldn't be seen as refinements of nested pairs because the row projection operators don't exist.
 %Labeled products are not refinements of binary products because the label projection operator is not defined for binary products.%Regular string types could also support an operation that the internal string type would not (e.g.
@@
 canonical forms) \cite{conf/popl/DelawareOS13}.
-The authors, along with Chlipala  \cite{Chlipala10}, suggest  proof automation. This is fundamentally a heuristic approach.
-
-In contrast, with @$\lambda$ fragment providers need not provide the semantics with mechanized proofs to benefit from modular reasoning principles.
+The authors, along with Chlipala  \cite{Chlipala10}, suggest  proof automation as a heuristic solution to the problem.
+
+In contrast, in @$\lambda$, fragment providers need not provide the semantics with mechanized proofs to benefit from modular reasoning principles.
 Instead, under a fixed tycon context, the calculus can be reasoned about like a very small type-directed compiler \cite{tarditi+:til-OLD,conf/pldi/Chlipala07}.

### 2014-11-06 att-pldi15/att-pldi15.tex
Minor edits / refactoring

@@
 
-\title{Modular Type Constructors}
-\subtitle{Conservatively Composing Typed Language Fragments}
+\title{Modular Composing Typed Language Fragments}
+%\subtitle{Modular Type Constructors}
 
@@
 \item
-\textbf{Specialized Fragments:} Fragments that track specialized static invariants are also frequently introduced in dialects, e.g.
+\textbf{Specialized Fragments:} Fragments providing more specialized operators are also frequently introduced in dialects, e.g.
 for distributed programming \cite{Murphy:2007:TDP:1793574.1793585}, reactive programming \cite{mandel2005reactiveml}, databases \cite{Ohori:2011:MSM:2034773.2034815},  units of measure \cite{conf/cefp/Kennedy09} and regular string sanitation \cite{sanitation-psp14}, amongst many others.
@@
 
-We call a tycon associated with opcon definitions a \emph{modular tycon} because all translation invariants maintained by the opcons associated with a tycon in any ``closed world'', e.g.
+We can call a tycon associated with opcons like this a \emph{modular tycon} because all translation invariants maintained by the opcons associated with a tycon in any ``closed world'', e.g.
 the regular string invariant just mentioned under the tycon context defining only $\tcvar{rstr}$, are necessarily maintained in any further extended tycon context, i.e.
-in the ``open world'', due to a simple static check that maintains \emph{translation independence} between tycons using type abstraction in the IL, the same fundamental mechanism underlying representation independence in ML-style module systems.
+in the ``open world'', due to a simple static check that maintains \emph{translation independence} between tycons using type abstraction in the IL, the same fundamental principle underlying representation independence in ML-style module systems.
 As in ML, mechanized specifications and proofs are not needed: the modularity guarantee is for all invariants.%, which provides a form of type-level computation.
@@
 Secondly, it will  give us a clean way to reuse the generalized intro form $\eintro{\st}{\es}$ and its desugarings at many types \cite{TSLs}.
-For example, regular string types can use standard string literal syntax and variants of record types can share   forms like $\{\conclbl{lbl1}=e_1, \conclbl{lbl2}=e_2\}$.
+For example, regular string types will use standard string literal syntax and variants of record types can share   forms like $\{\conclbl{lbl1}=e_1, \conclbl{lbl2}=e_2\}$.
 
@@
 The portion of the SL covered by the first row of kinds and static terms, some of which are elided for concision, forms an entirely standard total functional programming language consisting of total functions, polymorphic kinds, inductive kinds, and products and sums  \cite{pfpl}.
-The reader can consider these as forming a total subset of ML \cite{harper1997programming} or a simply-typed subset of Coq \todo{citation}and we will assume standard conveniences in examples (e.g.
+The reader can consider these as forming a total subset of ML \cite{harper1997programming} or a simply-typed subset of Coq \cite{Coq:manual} and we will assume standard conveniences in examples (e.g.
 let bindings and inference of type parameters) for concision.
@@
 \end{figure*}
+
+\noindent
+External types, or simply \emph{types}, are static values of kind $\kty$.
+We write $\istype{\st}{\Phi}$  iff $\sofkn{\emptyset}{\emptyset}{\Phi}{0}{\st}{\kty}$ and $\sval{\st}{\cdot;\emptyset;\Phi}$.
+The introductory form for kind $\kty$ is $\sty{c}{\st}$, where $c$ is a \emph{tycon} and $\st$ is the \emph{type index}.
+The dynamics are simple: the index is eagerly normalized and errors propagate  (the supplement gives the full set of rules for this paper).
+
+The syntax  for tycons given in Figure \ref{syntax-TC} specifies that $c$ can have one of three forms, so there are three corresponding kinding rules for types, shown in Figure \ref{fig:types}.
+
+\paragraph{Function Types} The rule \rulename{k-parr-ty} specifies that the type index of partial function types must be a pair of types.
+We thus say that $\rightharpoonup$ has \emph{index kind} $\kty \times \kty$.
+In practice, we would introduce a desugaring from $\st_1 \rightharpoonup \st_2$ to $\sty{\rightharpoonup}{(\st_1, \st_2)}$.
+%Here, we will benefit by treating it uniformly.
+
+\paragraph{Extension Types} For types constructed by an \emph{extension tycon}, written in small caps, $\tc$, the rule \rulename{k-ext-ty} extracts the index kind of $\tc$ from the tycon context, $\Phi$.
+
 \begin{figure}[t]
@@
 
-\noindent
-External types, or simply \emph{types}, are static values of kind $\kty$.
-We write $\istype{\st}{\Phi}$  iff $\sofkn{\emptyset}{\emptyset}{\Phi}{0}{\st}{\kty}$ and $\sval{\st}{\cdot;\emptyset;\Phi}$.
-The introductory form for kind $\kty$ is $\sty{c}{\st}$, where $c$ is a \emph{tycon} and $\st$ is the \emph{type index}.
-The dynamics are simple: the index is eagerly normalized and errors propagate  (the supplement gives the full set of rules for this paper).
-
-The syntax  for tycons given in Figure \ref{syntax-TC} specifies that $c$ can have one of three forms, so there are three corresponding kinding rules for types, shown in Figure \ref{fig:types}.
-
-\paragraph{Function Types} The rule \rulename{k-parr-ty} specifies that the type index of partial function types must be a pair of types.
-We thus say that $\rightharpoonup$ has \emph{index kind} $\kty \times \kty$.
-In practice, we would introduce a desugaring from $\st_1 \rightharpoonup \st_2$ to $\sty{\rightharpoonup}{(\st_1, \st_2)}$.
-%Here, we will benefit by treating it uniformly.
-
-\paragraph{Extension Types} For types constructed by an \emph{extension tycon}, written in small caps, $\tc$, the rule \rulename{k-ext-ty} extracts the index kind of $\tc$ from the tycon context, $\Phi$.
-
 For example,  $\tcvar{rstr}$ specifies the index kind $\krx$, which classifies static regular expression patterns (defined as an inductive sum kind in the usual way).
@@
 
-Two tycon contexts, each defining just one of the tycons just mentioned, are shown in Figure \ref{fig:example-tycons}.\footnote{In examples, we omit leading $\emptyset$, used as the base case for finite mappings, and $\cdot$, used as the base case for finite sequences, for concision.} We write the tycon context containing both  $\Phi_\text{rstr}\Phi_\text{lprod}$.
+Two tycon contexts, each defining just one of the tycons just mentioned, are shown in Figure \ref{fig:example-tycons}.\footnote{In examples, we omit leading $\emptyset$, used as the base case for finite mappings, and $\cdot$, used as the base case for finite sequences, for concision.}
 The syntax of tycon contexts in Figure \ref{syntax-TC} specifies that they are simply lists of tycon definitions, $\tcdef{\tc}{\psi}{\theta}$.
@@
 The tycon context well-definedness judgement, written $\vdash \Phi$, requires that all tycon names are unique and performs three additional checks, shown in Figure \ref{fig:tycon-context-well-definedness} and described below (we omit the trivial base case, (tcc-emp)).
+We write tycon context concatenation as, e.g., $\Phi_\text{rstr}\Phi_\text{lprod}$.
 
@@
 \begin{figure}\small
-\hspace{-10px}\begin{tabular}{ll}
+\hspace{-10px}\begin{tabular}{l|l}
 $\begin{array}{l}\Phi_\text{rstr} := \tcdef{\tcvar{rstr}}{\psi_\text{rstr}}{\\
-\quad\tcstruct{\stx{rstr/trans}}{\\
+\quad\tcstruct{\stx{rstr/schema}}{\\
 \quad\tcstructn{\stx{rstr/intro}}};\\
@@
 \quad\cdots}\end{array}$ & $\begin{array}{l}\Phi_\text{lprod} := \tcdef{\tcvar{lprod}}{\psi_\text{lprod}}{\\
-\quad\tcstruct{\stx{lprod/trans}}{\\
+\quad\tcstruct{\stx{lprod/schema}}{\\
 \quad\tcstructn{\stx{lprod/intro}}};\\
@@
 $\psi_\text{lprod} := \tcsig{\klist{\klbl\times\kty}}{\introsig{\klist{\klbl}}; \opname{\#}[\klbl]; \opname{conc}[\kunit]; ...}$
-\caption{Example tycon contexts and signatures.}
+\caption{Example tycon definitions and signatures.}
 \label{fig:example-tycons}\vspace{-8px}
@@
 those for which semantically equivalent values are syntactically equal.
-We define these by the judgement $\keq{\kDelta}{\kappa}$ (see supplement).
+Equality kinds are defined by the judgement $\keq{\kDelta}{\kappa}$ (see supplement).
 Arrow kinds are not equality kinds, so type indices cannot contain static functions.
-Equality kinds are exactly analagous to equality types as found in Standard ML \cite{Tofte:89:TheDefinitionOfStandardML}.
+Equality kinds are exactly analagous to equality types as in Standard ML \cite{Tofte:89:TheDefinitionOfStandardML}.
 
@@
 Applying this translation schema to the index of $\stx{paper}$, for example, produces  $\stx{paper/trans} := \sqity{\qity_\text{paper/trans}}$ where $\small\qity_\text{paper/trans} := \srep{\stx{title}}\times(\srep{\stx{conf}}\times \iunit)$.
-Note that our schema did not remove the trailing unit type for simplicity (and to again emphasize that this choice is local).
-It may be helpful to derive that $\stx{paper/trans} :: \kity$ by the rules above.%For simplicity, and to make the point that the choice of representation has only tycon-local implications, our translation schema does not optimize away
+Note that our schema did not remove the trailing unit type for simplicity.% (and to again emphasize that this choice is local).
+%It may be helpful to derive that $\stx{paper/trans} :: \kity$ by the rules above.%For simplicity, and to make the point that the choice of representation has only tycon-local implications, our translation schema does not optimize away
 
@@
   \stx{def}(\sttyidx)(\stmidx)(\stx{args}) \Downarrow_{\es;\Upsilon;\Phi} \sqitm{\qitm}\\\\
-  \trvalidate{\es;\Upsilon;\Phi}{\tc}{\qitm}{\srep{\sty{\tc}{\sttyidx}}}{\iota}
+  \trvalidate{\es;\Upsilon;\Phi}{\tc}{\qitm}{{\sty{\tc}{\sttyidx}}}{\iota}
   %\validate{1}{2{3}{4}{5}
@@
   \stx{def}(\sttyidx)(\stmidx)(\stx{args}) \Downarrow_{(e_\text{targ}; \es);\Upsilon;\Phi} (\st, \sqitm{\qitm})\\\\
-  \trvalidate{(e_\text{targ}; \es);\Upsilon;\Phi}{\tc}{\qitm}{\srep{\st}}{\iota}
+  \trvalidate{(e_\text{targ}; \es);\Upsilon;\Phi}{\tc}{\qitm}{{\st}}{\iota}
   %\snorm{\stx{def}~\sttyidx~\stmidx~\stx{args}}{\emptyset}{\memG_0}{{\rightharpoonup}, \tc; \Upsilon; \Phi}{\sqitm{\itm_\text{abs}}}{\memD}{\memG}\\\\
@@
     \stx{def}(\stx{args}) \Downarrow_{\es; \Upsilon; \Phi} \sqitm{\qitm}\\\\
-    \trvalidate{\es;\Upsilon;\Phi}{\keyw{other}[m;\kappa]}{\qitm}{\srep{\sty{\keyw{other}[m;\kappa]}{\sttyidx}}}{\iota}
+    \trvalidate{\es;\Upsilon;\Phi}{\keyw{other}[m;\kappa]}{\qitm}{{\sty{\keyw{other}[m;\kappa]}{\sttyidx}}}{\iota}
 }{
@@
   \stx{def}(\stx{args}) \Downarrow_{(e_\text{targ}; \es);\Upsilon;\Phi} (\st, \sqitm{\qitm})\\\\
-  \trvalidate{(e_\text{targ}; \es);\Upsilon;\Phi}{\keyw{other}[m; \kappa]}{\qitm}{\srep{\st}}{\iota}
+  \trvalidate{(e_\text{targ}; \es);\Upsilon;\Phi}{\keyw{other}[m; \kappa]}{\qitm}{{\st}}{\iota}
   %\snorm{\stx{def}~\sttyidx~\stmidx~\stx{args}}{\emptyset}{\memG_0}{{\rightharpoonup}, \tc; \Upsilon; \Phi}{\sqitm{\itm_\text{abs}}}{\memD}{\memG}\\\\
@@
 \noindent
-The judgement $\trvalidate{\argEnv}{c}{\qitm}{\srep{\st}}{\itm}$, defined by a single rule in Figure \ref{fig:translation-validation} and appearing as the final premise of (ana-intro), can be pronounced ``translational internal term $\qitm$ generated by an opcon associated with $c$ under argument environment $\argEnv$ is a valid translation for an external term of type $\st$, so the actual translation $\iota$ is produced''.
-For example, $$\trvalidate{(title; \concstr{EXMPL 2015}); \Upsilon_\text{ex}; \Phi_\text{rstr}\Phi_\text{lprod}}{\tcvar{lprod}}{\qitm_\text{ex}}{\srep{\stx{paper}}}{\iota_\text{ex}}$$
+The judgement $\trvalidate{\argEnv}{c}{\qitm}{\st}{\itm}$, defined by a single rule in Figure \ref{fig:translation-validation} and appearing as the final premise of (ana-intro) and the other rules described below, is pronounced ``translational internal term $\qitm$ generated by an opcon associated with $c$ under argument environment $\argEnv$ for an operation with type $\st$ is valid, so translation $\iota$ is produced''.
+For example, $$\trvalidate{(title; \concstr{EXMPL 2015}); \Upsilon_\text{ex}; \Phi_\text{rstr}\Phi_\text{lprod}}{\tcvar{lprod}}{\qitm_\text{ex}}{{\stx{paper}}}{\iota_\text{ex}}$$
 
@@
 \begin{figure}
-\small\fbox{$\trvalidate{\argEnv}{c}{\qitm}{\srep{\st}}{\iota\moutput}$}\vspace{-5px}
+\small\fbox{$\trvalidate{\argEnv}{c}{\qitm}{\st}{\iota\moutput}$}\vspace{-5px}
 \begin{mathpar}\small
@@
 }{
-\trvalidate{\es;\Upsilon;\Phi}{c}{\qitm}{\srep{\st}}{[\delta][\gamma]\iota_\text{abs}}
+\trvalidate{\es;\Upsilon;\Phi}{c}{\qitm}{\st}{[\delta][\gamma]\iota_\text{abs}}
 }\end{mathpar}\vspace{-9px}
@@
 \ref{sec:selective-type-translation-abstraction}.
-Like type translation stores, each term translation store induces an internal term substitution,  $\gamma ::= \emptyset ~|~ \gamma, \iota/x$, and corresponding internal typing context $\Gamma$ by the judgement $\memG \leadsto \gamma : \Gamma$, appearing as the next premise.
+Like type translation stores, each term translation store induces an internal term substitution,  $\gamma ::= \emptyset ~|~ \gamma, \iota/x$, and corresponding internal typing context $\Gamma$ by the judgement $\memG \leadsto \gamma : \Gamma$, appearing as the fourth premise.
 Here, $\small
@@
 \[\small(\alpha_0, \alpha_1)~(x_0 : \alpha_0, x_1 : \alpha_1) \vdash (x_0, (x_1, ())) : \alpha_0 \times (\alpha_1 \times \kunit)\]
-So, the translation of the labeled product $e_\text{ex}$ generated by $\tcvar{lprod}$ is checked with the references to term and type translations of regular strings replaced by variables and type variables, respectively.
-But because our  definition treated arguments parametrically, the check succeeds.% We  describe  an ill-behaved operator in Sec.
+The result is that the translation of the labeled product $e_\text{ex}$ generated by $\tcvar{lprod}$ is checked with the references to term and type translations of regular strings replaced by variables and type variables, respectively.
+But because our  definition treated arguments parametrically, the check succeeds.
+Had we attempted to smuggle in a regular string that violated the regular string invariant, e.g.
+generating $\qitm_\text{bad} := (\concstr{}, (\concstr{}, ())$, the check fails (even though $(\concstr{}, (\concstr{}, ()) : \keyw{str} \times \keyw{str} \times \kunit$).% We  describe  an ill-behaved operator in Sec.
 \ref{metatheory}.
@@
 The rule (ocstruct-targ) in Figure \ref{ocstruct} ensures that no tycon defines an opcon twice and that the opcon definitions are well-kinded.
-For example, $\stx{rstr/conc} :=$
-\[\small
-\begin{array}{l}
+For example, the definition of $\tcvar{rstr}$'s  $\opname{conc}$ opcon is shown in Figure \ref{fig:example-conc}.
+
+\begin{figure}[t]
+$\small\begin{array}{l}
 \keyw{syn}~\opname{conc}=\slam{\krx}{\svar{tyidx}}{
@@
 \end{array}
-\]
+$
+\caption{A targeted opcon definition, $\stx{rstr/conc}$.}
+\label{fig:example-conc}
+\vspace{-8px}
+\end{figure}
 The helper function $\svar{arity2}$ checks that two arguments, including the target argument, were provided.
@@
 We case analyze the second against $\tcvar{rstr}$, extracting its index regular expression if so and raising an error if not.
-We then synthesize the resulting regular string type, using the helper function $\svar{rxconcat} :: \krx \rightarrow \krx \rightarrow \krx$ which generates the synthesized type index by concatenating the indices of the argument's types, and finally the translation, using an internal helper function $sconcat : \keyw{str} \rightarrow \keyw{str} \rightarrow \keyw{str}$, the translational term for which we assume has been substituted in directly.
+We then synthesize the resulting regular string type, using the helper function $\svar{rxconcat} :: \krx \rightarrow \krx \rightarrow \krx$ which generates the synthesized type's  index by concatenating the indices of the argument's types, and finally the translation, using an internal helper function $sconcat : \keyw{str} \rightarrow \keyw{str} \rightarrow \keyw{str}$, the translational term for which we assume has been substituted in directly.
 
@@
 \subsection{Operations Over Other Types}
-The rule (ana-intro-other) is used to introduce terms of a type constructed by an ``other'' tycon.
-The term index, rather than the tycon context, directly specifies the static function that maps the arguments to a translation.
-In all other respects, it is analagous.
-It is used as a technical device in Sec.
+The rules (ana-intro-other) and (syn-targ-other) are used to introduce and simulate other operations on terms of a type constructed by an ``other'' tycon.
+In both cases, the term index, rather than the tycon context, directly specifies the static function that maps the arguments to a translation.
+%In all other respects, they are familiar.
+%They are used as a technical device in Sec.
 \ref{metatheory}.
 
-Like (ana-intro-other), rule (syn-targ-other) is used when the target synthesizes an ``other'' type.
+%Like (ana-intro-other), rule (syn-targ-other) is used when the target synthesizes an ``other'' type.
 The mapping from the arguments to a type and translation is again given directly in the term index (the op name is ignored).
@@
 canonical forms) \cite{conf/popl/DelawareOS13}.
-The authors, along with Chlipala  \cite{Chlipala10}, suggest  proof automation as a heuristic solution to the problem.
-
-In contrast, in @$\lambda$, fragment providers need not provide the semantics with mechanized proofs to benefit from modular reasoning principles.
+Several authors, e.g. Chlipala  \cite{Chlipala10}, suggest  proof automation as a heuristic solution to the problem.
+
+In contrast, in @$\lambda$, fragment providers need not provide the semantics with mechanized proofs to benefit from rigorous modular reasoning principles.
 Instead, under a fixed tycon context, the calculus can be reasoned about like a very small type-directed compiler \cite{tarditi+:til-OLD,conf/pldi/Chlipala07}.

### 2014-11-16 att-pldi15/att-pldi15-supplement.tex
final PLDI15 submission

--- 2014-11-02 att-pldi15/att-pldi15-supplement.tex
@@
 \usepackage{mathpartir}
+\usepackage{enumerate}
 \usepackage[usenames,dvipsnames,svgnames,table]{xcolor}
-
+\usepackage{tabularx}
 \input{macros-catlam}
+\usepackage{morefloats}
 
@@
 \newtheorem{tyconinvariant}{Tycon Invariant}
+\newtheorem{assumption}{Assumption}
 \newenvironment{proof-sketch}{\noindent{\emph{Proof Sketch.}}}{\qed}
@@
 
+\newcommand{\pfcase}[1]{\textbf{Case}~#1. }
+\newcommand{\pfcases}[1]{\textbf{Cases}~#1. }
+
+\usepackage{hyperref}
+\hypersetup{
+    colorlinks,
+    citecolor=black,
+    filecolor=black,
+    linkcolor=black,
+    urlcolor=black
+}
+
 \begin{document}
-
-\conferenceinfo{-}{-} 
-\copyrightyear{-} 
-\copyrightdata{[to be supplied]}
 
@@
 
+\tableofcontents 
+\listoffigures
+
+\newpage
+
 \section{Internal Language}
-We assume the statics of the IL are specified in the standard way by judgements for  type formation {$\iDelta \vdash \tau$}, typing context formation { $\iDelta \vdash \iGamma$} and type assignment {$\iDelta~\iGamma \vdash \iota : \tau\moutput$}.
+\subsection{Semantics}
+We assume an internal language where the statics are specified in the standard way by judgements for  type formation {$\iDelta \vdash \tau$}, typing context formation { $\iDelta \vdash \iGamma$} and type assignment {$\iDelta~\iGamma \vdash \iota : \tau\moutput$}.
 %The typing and type formation contexts obey standard structural properties (i.e. weakening, exchange and contraction). 
@@
 
-We use {$\mathcal{L}\{{\rightharpoonup}\,{\forall}\,{\mu}\,{\iunit}\,{\times}\,{+}\}$}, the syntax for which is shown in Figure \ref{syntax-IL}, as representative of any intermediate language for a typed functional language.
+Our particular choice in the paper is {$\mathcal{L}\{{\rightharpoonup}\,{\forall}\,{\mu}\,{\iunit}\,{\times}\,{+}\}$}, the syntax for which is shown in Figure \ref{syntax-IL}, as representative of any intermediate language for a typed functional language.
 In fact, our intention is not to prescribe a particular choice of IL, so we will here only review the key metatheoretic properties that the IL must possess.
@@
 
-\begin{lemma}[Internal Type Assignment]
+\begin{assumption}[Internal Type Assignment]
 If $\Delta \vdash \Gamma$ and $\Delta~\Gamma \vdash \iota : \tau$ then $\Delta \vdash \tau$.
-\end{lemma}
-
-Internal type safety follows the standard methodology of proving preservation and progress lemmas.
-
-\begin{lemma}[Internal Progress]
+\end{assumption}
+
+Internal type safety follows the standard methodology: preservation and progress lemmas.
+
+\begin{assumption}[Internal Progress]
 If $\emptyset~\emptyset \vdash \iota : \tau$ then either $\iota~\texttt{val}$ or $\iota \mapsto \iota'$.
+\end{assumption}
+\begin{assumption}[Internal Preservation]
+If $\emptyset~\emptyset \vdash \iota : \tau$ and $\iota \mapsto \iota'$ then $\emptyset~\emptyset \vdash \iota' : \tau$. 
+\end{assumption}
+
+We assume that typing and type formation contexts, $\Delta$ and $\Gamma$, are identified up to exchange and contraction and the internal semantics obeys weakening.
+We assume that the names of variables and type variables are unimportant, so that $\alpha$-equivalent terms, types and contexts are identified implicitly throughout this work.
+
+\subsection{Explicit Substitutions}
+We also assume substitution is defined in the usual way.
+We define explicit $n$-ary substitutions for type variables, $\delta$, and term variables, $\gamma$, also in the standard way, writing $[\delta]\Gamma$, $[\delta]\tau$ and $[\delta]\iota$ for application of type substitutions and $[\gamma]\iota$ for application of term substitutions.
+We need the definitions of substitution validity shown in Figures \ref{fig:internal-type-substitution-validity} and \ref{fig:internal-term-substitution-validity} and assume the following lemmas.
+
+\begin{assumption}[Internal Term Substitutions] 
+\label{assumption:internal-term-substituions}
+If $\Delta \vdash \Gamma$ and $\Delta \vdash \Gamma'$ and $\Delta~\Gamma \vdash \gamma : \Gamma'$ and $\Delta~\Gamma\Gamma' \vdash \iota : \tau$ then $\Delta~\Gamma \vdash [\gamma]\iota : \tau$.
+\end{assumption}
+
+\begin{assumption}[Internal Type Substitution on Types]
+\label{lemma:internal-type-substitution-on-types} If $\Delta \vdash \delta : \Delta'$ and $\Delta\Delta' \vdash \tau$ then $\Delta \vdash [\delta]\tau$.\end{assumption}
+
+\begin{assumption}[Internal Type Substitutions on Typing Contexts] 
+\label{lemma:internal-type-substitutions-on-typing-contexts}
+If $\Delta \vdash \delta : \Delta'$ and $\Delta\Delta' \vdash \Gamma$ then $\Delta \vdash [\delta]\Gamma$.
+\end{assumption}
+
+\begin{assumption}[Internal Type Substitutions on Terms] 
+\label{lemma:internal-type-substitutions-on-terms}
+If $\Delta \vdash \delta : \Delta'$ and $\Delta\Delta' \vdash \tau$ and $\Delta\Delta' \vdash \Gamma$ and $\Delta\Delta'~\Gamma \vdash \iota : \tau$ then $\Delta~[\delta]\Gamma \vdash [\delta]\iota : [\delta]\tau$.
+\end{assumption}
+
+\newpage
+\section{Tycon Contexts}
+Tycon contexts, $\Phi$, are semantically defined as ordered mappings from tycon names $\tc$ to tycon definitions.
+As with all ordered mappings, we implicitly assume that all names are unique and that certain operations, e.g.
+$\text{dom}(\Phi)$, which extracts a set of tycon names, and $\tcdef{\tc}{\psi}{\theta} \in \Phi$, which checks whether a tycon definition is in $\Phi$, and $\Phi\Phi'$, which concatenates two tycon contexts with disjoint domains, are available without defining them explicitly.
+
+\subsection{Validity}
+
+The tycon context validity judgement $\vdash \Phi$ is specified in Figure \ref{fig:tycon-ctxs}. The kinding and kind formation rules will be given in the next section.
+
+\begin{lemma}[Tycon Context Validity]
+\label{lemma:tycon-context-validity}
+If (a) $\vdash \Phi$ and (b) $\tcdef{\tc}{\tcsig{\ktyidx}{\chi}}{\tcstruct{\stx{schema}}{\omega}} \in \Phi$ then there exists a prefix $\Phi'$ of $\Phi$ such that (i) $\vdash \Phi'$ and (ii) $\vdash \Phi', \tcdef{\tc}{\tcsig{\ktyidx}{\chi}}{\tcstruct{\stx{schema}}{\omega}}$ and (iii) $\keq{}{\ktyidx}$ and (iv) $\sofkz{\emptyset}{\emptyset}{\Phi'}{\stx{schema}}{\ktyidx \rightarrow \kity}$ and (v) $\vdash_{\Phi',  \tcdef{\tc}{\tcsig{\ktyidx}{\chi}}{\tcstruct{\stx{schema}}{\omega}}} \omega \sim \tcsig{\ktyidx}{\chi}$.
+\end{lemma}
+\begin{proof} Rule induction on (a) and (b).
+
+\pfcase{(tcc-emp)} Does not apply by (b).
+
+\pfcase{(tcc-ext)} We have two cases by (b):\begin{quote}
+\pfcase{$\Phi=\Phi', \tcdef{\tc}{\tcsig{\ktyidx}{\chi}}{\tcstruct{\stx{schema}}{\omega}}$} In this case, we pick $\Phi'$ and have our conclusions:
+\begin{enumerate}
+\item[(i)] $\vdash \Phi'$ (premise 1 of (tcc-ext))
+\item[(ii)] $\vdash \Phi', \tcdef{\tc}{\tcsig{\ktyidx}{\chi}}{\tcstruct{\stx{schema}}{\omega}}$ (by assumption (a))
+\item[(iii)] $\keq{}{\ktyidx}$ (premise 2 of (tcc-ext))
+\item[(iv)] $\sofkz{\emptyset}{\emptyset}{\Phi'}{\stx{schema}}{\ktyidx \rightarrow \kity}$ (premise 3 of (tcc-ext))
+\item[(v)] $\vdash_{\Phi', \tcdef{\tc}{\tcsig{\ktyidx}{\chi}}{\tcstruct{\stx{schema}}{\omega}}} \omega \sim \tcsig{\ktyidx}{\chi}$ (premise 4 of (tcc-ext))
+\end{enumerate}
+
+\pfcase{$\Phi=\Phi', \tcdef{\tc'}{\psi'}{\theta'}$ for $\tc\neq\tc'$ and (1) $\tcdef{\tc}{\tcsig{\ktyidx}{\chi}}{\tcstruct{\stx{schema}}{\omega}} \in \Phi'$} We have the following:
+
+\begin{enumerate}
+\item[(2)] $\vdash \Phi'$ (premise 1 of (tcc-ext))
+\item[] Apply IH to (2) and (1). There exists a prefix $\Phi''$ of $\Phi'$ such that 
+\item[(3)] $\vdash \Phi''$
+\item[(4)] $\vdash \Phi'', \tcdef{\tc}{\tcsig{\ktyidx}{\chi}}{\tcstruct{\stx{schema}}{\omega}}$
+\item[(5)] $\keq{}{\ktyidx}$ 
+\item[(6)] $\sofkz{\emptyset}{\emptyset}{\Phi''}{\stx{schema}}{\ktyidx \rightarrow \kity}$
+\item[(7)] $\vdash_{\Phi'',  \tcdef{\tc}{\tcsig{\ktyidx}{\chi}}{\tcstruct{\stx{schema}}{\omega}}} \omega \sim \tcsig{\ktyidx}{\chi}$
+\end{enumerate}
+
+By transitivity of prefixing, we have that $\Phi''$ is a prefix of $\Phi$. Picking $\Phi''$, conclusions (i-v) are (3-7), respectively.
+\end{quote}
+\end{proof}
+
+In the final premise of (tcc-ext), the opcon structure is checked against the tycon's signature by the rules in Figure \ref{ocstruct}.
+Opcon structures and opcon signatures are finite maps, so we implicitly assume that all operator names are unique and standard operations are defined as above.
+
+\begin{lemma}[Intro Opcon Existence and Validity]
+\label{lemma:intro-opcon-existence-and-validity}
+If $\vdash_\Phi \omega \sim \tcsig{\ktyidx}{\chi}$ then (i) $\keyw{intro}[\klitidx] \in \chi$ and (ii) $\emptyset \vdash \klitidx$ and (iii) $\keyw{ana~intro}=\stx{def}\in\omega$ and (iv) $\sofkz{\emptyset}{\emptyset}{\Phi}{\stx{def}}{\ktyidx \rightarrow \klitidx \rightarrow \kargs \rightarrow \kitm}$.
+\end{lemma}
+\begin{proof}
+Rule induction on assumption.
+
+\pfcase{(ocstruct-intro)} We have that $\omega=\keyw{ana~intro}=\stx{def}$ and $\chi=\keyw{intro}[\klitidx]$.
+
+\begin{enumerate}[(i)]
+\item $\keyw{intro}[\klitidx] \in \keyw{intro}[\klitidx]$ (by definition)
+\item $\emptyset \vdash \klitidx$  (premise 1)
+\item  $\keyw{ana~intro}=\stx{def}\in\keyw{ana~intro}=\stx{def}$ (by definition)
+\item $\sofkz{\emptyset}{\emptyset}{\Phi}{\stx{def}}{\ktyidx \rightarrow \klitidx \rightarrow \kargs \rightarrow \kitm}$ (premise 2)
+\end{enumerate}
+
+\pfcase{(ocstruct-targ)} We have that $\omega=\tcstructc{\omega'}{op}{\stx{def}}$ and  $\chi=\chi'; \opsig{\opname{op}}{\klitidx'}$.
+\begin{enumerate}[(1)]
+\item $\vdash_\Phi \omega' \sim \tcsig{\ktyidx}{\chi}$ (premise 1)
+\item[] Apply IH to (1):
+\item $\introsig{\klitidx} \in \chi'$
+\item $\emptyset \vdash \klitidx$
+\item $\keyw{ana~intro}=\stx{def}\in\omega'$
+\item $\sofkz{\emptyset}{\emptyset}{\Phi}{\stx{def}}{\ktyidx \rightarrow \klitidx \rightarrow \kargs \rightarrow \kitm}$
+\end{enumerate}
+We can now conclude:
+\begin{enumerate}[(i)]
+\item $\introsig{\klitidx} \in \chi$ by extension of finite mappings on (2)
+\item $\emptyset \vdash \klitidx$ by (3)
+\item $\keyw{ana~intro}=\stx{def}\in\omega$ by extension of finite mappings on (4)
+\item $\sofkz{\emptyset}{\emptyset}{\Phi}{\stx{def}}{\ktyidx \rightarrow \klitidx \rightarrow \kargs \rightarrow \kitm}$ by (5).
+\end{enumerate}
+\end{proof}
+\begin{lemma}[Targeted Opcon Validity]
+\label{lemma:targeted-opcon-validity}
+If (a) $\vdash_\Phi \omega \sim \tcsig{\ktyidx}{\chi}$ and (b) $\keyw{syn}~\opname{op}=\stx{def} \in \omega$ then (i) $\opname{op}[\klitidx] \in \chi$ and (ii) $\emptyset \vdash \klitidx$ and (iii) $
+    \sofkn{\emptyset}{\emptyset}{\Phi}{0}{\stx{def}}{\ktyidx \rightarrow \klitidx \rightarrow \kargs \rightarrow (\kty \times \kitm)}$.
+\end{lemma}
+\begin{proof}
+Rule induction on assumption (a).
+
+\pfcase{(ocstruct-intro)} Does not apply by (b).
+
+\pfcase{(ocstruct-targ)} We have two cases by (b).\begin{quote}\pfcase{$\omega=\omega'; \keyw{syn}~\opname{op}=\stx{def}$ and $\chi=\chi', \opname{op}[{\klitidx}]$} Conclusion (i) follows by definition and conclusions (ii) and (iii) are premises 2 and 3 of (ocstruct-targ).
+
+\pfcase{$\omega=\omega'; \keyw{syn}~\opname{op}'=\stx{def}'$ such that $\opname{op}\neq\opname{op'}$ and (1) $\keyw{syn}~\opname{op}=\stx{def} \in \omega'$} In this case, we have 
+    \begin{enumerate}
+    \item[(2)] $\vdash_\Phi \omega' \sim \tcsig{\ktyidx}{\chi'}$ (premise 1 of (ocstruct-targ))
+    \end{enumerate}
+
+    Applying the IH to (2) and (1), we have:
+    \begin{enumerate}[(1)]
+    \item[(3)] $\opname{op}[\klitidx] \in \chi'$
+    \item[(4)] $\emptyset \vdash \klitidx$
+    \item[(5)] $\sofkn{\emptyset}{\emptyset}{\Phi}{0}{\stx{def}}{\ktyidx \rightarrow \klitidx \rightarrow \kargs \rightarrow (\kty \times \kitm)}$
+    \end{enumerate}
+
+    We thus have conclusion (i) by extension of finite mappings on (3), and conclusions (ii) and (iii) are (4) and (5).
+\end{quote}
+\end{proof}
+
+\subsection{Signature Validity}
+
+For the purposes of the SL, only the type signatures are relevant.\footnote{In fact, only the type index kind is relevant, though in future work the remainder will likely be and it is cleaner to define complete signature checking.} It is thus useful to define judgements for checking tycon signature validity separately.
+The judgements $\vdash \psi$ for tycon signatures, $\vdash \chi$ for opcon signatures, and $\vdash \Phi~\mathtt{sigsok}$ for all signatures in a tycon context are in Figure \ref{fig:tycon-and-opcon-signature-well-formedness}.
+Kind and equality kind formation are covered in the next section.
+
+\begin{lemma}[Signatures Define Equality Type Index Kinds]
+\label{lemma:sigs-define-eq-type-index-kinds}
+If $\vdash \tcsig{\ktyidx}{\chi}$ then $\keq{}{\ktyidx}$.
+\end{lemma}
+\begin{proof} By rule induction on the assumption. The conclusion is the first premise.
+\end{proof}
+
+\begin{lemma}[Valid Opcon Structures Define Valid Opcon Signatures]
+\label{lemma:chi-ok}
+If $\vdash_\Phi \omega \sim \tcsig{\ktyidx}{\chi}$ then $\vdash \chi$.
+\end{lemma}
+\begin{proof} By rule induction on the assumption.
+
+\pfcase{(ocstruct-intro)} We have that $\chi=\introsig{\klitidx}$ and (1) $\emptyset \vdash \ktyidx$. Apply (ocsig-intro-ok) to (1).
+
+\pfcase{(ocstruct-targ-ok)} We have that $\chi=\chi', \opsig{\opname{op}}{\klitidx}$ and (1) $\vdash_\Phi \omega \sim \tcsig{\klitidx}{\chi'}$ and (2) $\emptyset \vdash \klitidx$.
+Applying the IH to (1), we have (3) $\vdash \chi'$.
+Apply (ocsig-targ-ok) to (3) and (2).
+\end{proof}
+
+\begin{lemma}[Well-Defined Contexts Have Well-Formed Signatures]
+\label{lemma:well-defined-contexts-have-well-formed-signatures}
+If $\vdash \Phi$ then $\vdash \Phi~\mathtt{sigsok}$.
+\end{lemma}
+\begin{proof} By rule induction on the assumption.
+
+\pfcase{(tcc-emp)} Apply (tcc-sigsok-emp).
+
+\pfcase{(tcc-ext)} We have that $\Phi=\Phi', \tcdef{\tc}{\tcsig{\ktyidx}{\chi}}{\theta}$ and (1) $\vdash \Phi'$ and (2) $\keq{}{\ktyidx}$ and (3) $\vdash_\Phi \omega \sim \tcsig{\ktyidx}{\chi}$.
+
+By the IH on (1), we have (4) $\vdash \Phi'~\mathtt{sigsok}$.
+
+By Lemma \ref{lemma:chi-ok} on (3) we have (5) $\vdash \chi$.
+
+By (tcsig-ok) on (2) and (5) we have (6) $\vdash \tcsig{\ktyidx}{\chi}$.
+
+By (tcc-sigsok-ext) on (4) and (6) we have our conclusion.
+\end{proof}
+
+\begin{lemma}[Tycons Have Valid Signatures]
+\label{lemma:tycons-have-valid-signatures}
+If (a) $\tcdef{\tc}{\psi}{\theta} \in \Phi$ and (b) $\vdash \Phi~\mathtt{sigsok}$ then $\vdash \psi$.
+\end{lemma}
+\begin{proof} By rule induction on (b) and (a).
+
+\pfcase{(tcc-sigsok-emp)} Does not apply by (a).
+
+\pfcase{(tcc-sigsok-ext)} We have two possible cases by (a):
+\begin{quote}
+\pfcase{$\Phi=\Phi', \tcdef{\tc}{\psi}{\theta}$} Then the conclusion is premise 2 of (b).
+
+\pfcase{$\Phi=\Phi', \tcdef{\tc'}{\psi'}{\theta'}$ for $\tc\neq\tc'$ and (1) $\tcdef{\tc}{\psi}{\theta} \in \Phi'$} We have (2) $\tccsigok{\Phi'}$ by premise 1 of (b). Apply the IH to (1) and (2).
+\end{quote}
+\end{proof}
+
+\begin{lemma}[Tycon Context Signature Validity Respects Extension]
+If (a) $\tccsigok{\Phi}$ and (b) $\tccsigok{\Phi'}$ and (c) $\text{dom}(\Phi) \cap \text{dom}(\Phi') = \emptyset$ then $\tccsigok{\Phi\Phi'}$.
+\end{lemma}
+\begin{proof} Rule induction on (b).
+
+\pfcase{(tcc-sigsok-emp)} Here, $\Phi'=\emptyset$ so $\Phi\Phi'=\Phi$ and the conclusion follows by (a).
+
+\pfcase{(tcc-sigsok-ext)} Here, $\Phi'=\Phi'', \tcdef{\tc}{\tcsig{\ktyidx}{\chi}}{\tcstruct{\stx{schema}}{\omega}}$ and (1) $\tccsigok{\Phi''}$ and (2) $\vdash \psi$.
+
+$\Phi\Phi''$ is well-formed because prefixing respects disjointness and (c).
+
+Applying the IH on (a) and (1), we have that (3) $\tccsigok{\Phi\Phi''}$.
+
+By (tcc-sigsok-ext) on (3) and (2) we have our conclusion.
+\end{proof}
+\newpage
+
+% \itm & ::= & \multicolumn{1}{l}{x ~|~ \ilam{\ity}{x}{\itm} ~|~ \iap{\itm}{\itm} ~|~ \ifix{\ity}{x}{\itm} ~|~ \iLam{\alpha}{\itm} ~|~ \iAp{\ity}{\itm}}\\
+% & ~|~ & \multicolumn{1}{l}{\ifold{t}{\ity}{\itm} ~|~ \iunfold{\itm} ~|~ \itriv ~|~ \ipair{\itm}{\itm} ~|~ \ifst{\itm} ~|~ \isnd{\itm}}\\
+% & ~|~ & \iinl{\ity}{\itm} ~|~ \iinr{\ity}{\itm} ~|~ \icase{\itm}{x}{\itm}{x}{\itm}\\
+
+\section{Static Language}
+\subsection{Syntax}
+
+In the paper, we described the static language (SL) as consisting of a total functional core plus three additional kinds.
+That core is derived directly from \cite{pfpl}.
+Since the form of the judgements are not technically identical, we will re-state the rules below, but we will only sketch the corresponding cases in the proofs for concision (the core ignores the tycon context, argument bound $n$, etc.).
+
+The only change from the paper here is that we define the SL as having a built in list kind, rather than the more general mechanism for inductive kinds hinted at in the paper (lists are the only inductive kind we strictly need).
+The details of inductive kinds are in \cite{pfpl}, but again would distract from our main point and aren't needed to prove our theorems.
+Another approach is to simply allow recursive kinds, which would make it possible to define static terms that diverge.
+This does not break our fundamental guarantees, only decidability of typechecking (which we do not go into in detail here, as it is a more practical concern and requires non-trivial proofs of weak normalization that would also distract from our main point).
+
+As with the IL, each choice of SL forms a dialect of @$\lambda$.
+We anticipate that the SL (essentially, a domain-specific language for implementing typed translation semantics), and the IL would need to evolve much more slowly than the EL (e.g.
+the GHC compiler's IL has remained  stable for many years), so dialect formation should be needed far more rarely % mechanized proof of our properties for the SL should begin with an existing language mechanization.
+
+\subsection{Kinds}\label{sec:kinds}
+\subsubsection{Kind Formation}
+
+The kind formation rules are straightforward, shown in Figure \ref{fig:kind-formation}.
+Kind formation contexts $\kDelta$ are defined as finite mappings and thus are identified up to contraction and exchange, and we can assume the domains are disjoint.
+We assume substitution is defined in the usual way, so that standard substitution lemmas hold.
+
+\begin{assumption}[Kind Variable Substitution - Kinds]
+\label{lemma:kind-variable-substitution-kinds}
+If $\kDelta, \kalpha \vdash \kappa'$ and $\kDelta \vdash \kappa$ then $\kDelta \vdash [\kappa/\kalpha]\kappa'$.
+\end{assumption}
+
+\begin{lemma}[Weakening of Kind Formation]
+\label{lemma:weakening-of-kind-formation}
+If $\kDelta \vdash \kappa$ then $\kDelta\kDelta' \vdash \kappa$.
+\end{lemma}
+\begin{proof} Rule induction on assumption.
+
+\pfcases{(kf-unit), (kf-ty), (kf-ity), (kf-itm)} These rules are defined for all $\kDelta$, so we re-apply the same rule.
+
+\pfcases{(kf-arrow), (kf-prod), (kf-sum)} Apply the IH to premise 1 to get (1) $\kDelta\kDelta' \vdash \kappa$.
+Apply the IH to premise 2 to get (2) $\kDelta\kDelta' \vdash \kappa_2$.
+Re-apply the same rule to (1) and (2).
+
+\pfcase{(kf-list)} Here, $\kappa=\klist{\kappa'}$. Apply the IH to premise 1 to get (1) $\kDelta\kDelta' \vdash \kappa'$. Apply (kf-list) to (1).
+
+\pfcase{(kf-alpha)} By extension of finite mappings, if $\kalpha \in \kDelta$ then (1) $\kalpha \in \kDelta\kDelta'$. Apply (kf-alpha) to (1).
+
+\pfcase{(kf-forall)} Here, $\kappa=\kforall{\kalpha}{\kappa'}$.
+Apply the IH to premise 1 to get (1) $(\kDelta, \kalpha)\kDelta' \vdash \kappa'$.
+By exchange, $(\kDelta, \kalpha)\kDelta'=\kDelta\kDelta', \kalpha$ and thus we can apply (kf-forall) to (1).
+
+\end{proof}
+
+\subsubsection{Equality Kinds}
+
+Equality kinds are included to avoid complicating type equality (and guarantee \emph{argument independence} of values, see below).
+Arrow kinds, polymorphic kinds and term quotation kinds (because they are argument-dependent, see below) are not equality kinds (though an approach for quantifying over only equality kinds could be introduced to allow polymorphic equality kinds, as Standard ML provides \cite{Tofte:89:TheDefinitionOfStandardML}).
+The rules are given in Figure \ref{fig:equality-kinds}.
+%We do not define an equational theory for the language at all, so this is merely meant to head off the concerns of those who might wish to develop one in the future.
+
+\begin{lemma}[Equality Kind Well-Formedness]
+\label{lemma:equality-kind-well-formedness}
+If $\keq{}{\kappa}$ then $\emptyset \vdash \kappa$. 
+\end{lemma}
+\begin{proof} Rule induction on the assumption.
+
+\pfcase{(keq-list)} We have $\kappa=\klist{\kappa'}$ and (1) $\keq{}{\kappa'}$. Apply IH to (1) to get that (2) $\emptyset \vdash \kappa'$. Apply (kf-list) to (2).
+
+\pfcase{(keq-unit)} We have $\kappa=\kunit$. Apply (kf-unit).
+
+\pfcase{(keq-prod)} We have $\kappa=\kprod{\kappa_1}{\kappa_2}$ and (1) $\keq{}{\kappa_1}$ and $\keq{}{\kappa_2}$.
+Apply IH to (1) to get (3) $\emptyset \vdash \kappa_1$.
+Apply IH to (2) to get (4) $\emptyset \vdash \kappa_2$.
+Apply (kf-prod) to (3) and (4).
+
+\pfcase{(keq-sum)} We have $\kappa=\ksum{\kappa_1}{\kappa_2}$ and (1) $\keq{}{\kappa_1}$ and $\keq{}{\kappa_2}$.
+Apply IH to (1) to get (3) $\emptyset \vdash \kappa_1$.
+Apply IH to (2) to get (4) $\emptyset \vdash \kappa_2$.
+Apply (kf-sum) to (3) and (4).
+
+\pfcase{(keq-ty)} We have $\kappa=\kty$. Apply (kf-ty).
+
+\pfcase{(keq-ity)} We have $\kappa=\kity$. Apply (kf-ity).
+\end{proof}
+
+\subsubsection{Kinding Contexts}
+Kinding contexts map static variables, written in bold, to kinds.
+The kinding context formation judgement checks these kinds against a kind formation context.
+Because both are finite mappings, we implicitly identify them up to exchange and contraction and the domain can be assumed disjoint implicitly.
+Like the IL, static terms and kinds are implicitly identified up to $\alpha$-renaming of kind and static term variables, and we assume substitution is defined in the usual way.
+
+\begin{lemma}[Kinding Context Lookup]
+\label{lemma:kinding-context-lookup} If (a) $\kDelta \vdash \kGamma$ and (b) $\sx :: \kappa \in \kGamma$ then $\kDelta \vdash \kappa$.\end{lemma}
+\begin{proof} Rule induction on (a).
+
+\pfcase{(kctx-emp)} Does not apply by (b).
+
+\pfcase{(kctx-ext)} We have that $\kGamma=\kGamma', \sx' :: \kappa'$ and (1) $\kDelta \vdash \kGamma'$ and (2) $\kDelta \vdash \kappa$. There are two cases by (b):
+\begin{quote}
+\pfcase{$\sx=\sx'$ and $\kappa=\kappa'$} We have our conclusion by (2).
+
+\pfcase{$\sx\neq\sx'$ and (3) $\sx :: \kappa \in \kGamma'$} Apply the IH to (1) and (3).
+\end{quote}
+\end{proof}
+
+\begin{lemma}[Weakening of Kinding Context Formation]
+\label{lemma:weakening-of-kinding-context-formation}
+If $\kDelta \vdash \kGamma$ then $\kDelta\kDelta' \vdash \kGamma$.
+\end{lemma}
+\begin{proof} Rule induction on assumption
+
+\pfcase{(kctx-emp)} Apply (kctx-emp).
+
+\pfcase{(kctx-ext)} We have $\kGamma=\kGamma', \sx :: \kappa$.
+Applying the IH to premise 1, we have (1) $\kDelta\kDelta' \vdash \kGamma'$.
+By Lemma \ref{lemma:weakening-of-kind-formation} on premise 2, we have (2) $\kDelta\kDelta' \vdash \kappa$.
+Apply (kctx-ext) to (1) and (2).
+\end{proof}
+
+\subsection{Kinding}
+The kinding rules are given in Figures \ref{fig:kinding-core}-\ref{fig:kinding-ana-syn}.
+Figure \ref{fig:kinding-core} describes kinding for the core and the rules are directly based on those in \cite{pfpl}, only tacking $\Phi$ and $n$ onto each rule, but never inspecting either one.
+For cases of our proofs where an existing proof script would be satisfactory by simply inserting these additional subscripts onto the kinding judgement, we will simply note this.
+Figure \ref{fig:kinding-types} describes kinding for types.
+Figures \ref{fig:kinding-qity-new} and \ref{fig:kinding-qity-shared} describes kinding for quoted translational internal types.
+Figures \ref{fig:kinding-qitm-new} and \ref{fig:kinding-qitm-shared} describe kinding for quoted translational internal terms.
+Only the rules for forms specific to quoted terms in Figures \ref{fig:kinding-qity-new} and \ref{fig:kinding-qitm-new} are interesting; the rules in Figures \ref{fig:kinding-qity-shared} and \ref{fig:kinding-qitm-shared} simply recurse generically over all sub-terms of shared form with $\iota$ or $\tau$.
+We will give only representative example cases for these shared forms in our proofs, rather than covering each form separately.
+Figure \ref{fig:kinding-ana-syn} describes kinding for the SL-EL interface.
+
+\begin{lemma}[Kind Assignment]
+If (a) $\kDelta \vdash \kGamma$ and (b) $\sofkX{\st}{\kappa}$ then $\kDelta \vdash \kappa$.
+\end{lemma}
+\begin{proof} By rule induction on (b). The cases for the rules in Figure \ref{fig:kinding-core} follow the usual script and are omitted.
+
+% \pfcase{(k-var)} $\st=\sx$ and (1) $\sx :: \kappa \in \kGamma$. Apply Lemma \ref{lemma:kinding-context-lookup} on (a) and (1).
+
+% \pfcase{(k-abs)} $\st=\slam{\kappa_1}{\sx}{\st'}$ and $\kappa=\kappa_1\rightarrow\kappa_2$ and (1) $\kDelta \vdash \kappa_1$ and (2) $\sofk{\kDelta}{\kGamma, \sx :: \kappa_1}{\Phi}{\st'}{\kappa_2}$.
+By IH on (a) and (2), we have (3) $\kDelta \vdash \kappa_2$.
+Apply (kf-arrow) to (1) and (3).
+
+% \pfcase{(k-ap)} $\st=\st_1(\st_2)$ and (1) $\sofkX{\st_1}{\karrow{\kappa_1}{\kappa_2}}$.
+By IH on (a) and (1), we have (2) $\kDelta \vdash \karrow{\kappa_1}{\kappa_2}$.
+By inversion of (kf-arrow), $\kDelta \vdash \kappa_2$.
+
+% \pfcase{(k-kabs)} $\st=\sLam{\kalpha}{\st'}$ and $\kappa=\kforall{\kalpha}{\kappa'}$ and (1) $\sofk{\kDelta, \kalpha}{\kGamma}{\Phi}{\st'}{\kappa}$.
+By IH to (a) and (1), we have (2) $\kDelta, \kalpha \vdash \kappa$.
+Apply (kf-forall) on (2).\todo{add context formation steps}
+
+% \pfcase{(k-kap)} $\st=\skap{\kappa'}{\st'}$ and (1) $\sofkX{\st'}{\kforall{\kalpha}{\kappa''}}$ and (2) $\kDelta \vdash \kappa'$ and $\kappa=[\kappa'/\kalpha]\kappa''$.
+By IH on (a) and (1), we have (2) $\kDelta \vdash \kforall{\kalpha}{\kappa''}$.
+By inversion of (kf-forall) on (2), we have (3) $\kDelta, \kalpha \vdash \kappa''$.
+Apply Lemma \ref{lemma:kind-variable-substitution-kinds} on (3) and (2).
+
+% \pfcase{(k-nil)} $\st=\keyw{nil}[\kappa']$ and $\kappa=\klist{\kappa'}$ and (1) $\kDelta \vdash \kappa'$. Apply (kf-list) on (1).
+
+% \pfcase{(k-cons)} $\st=\keyw{cons}(\st_1; \st_2)$ and $\kappa=\klist{\kappa'}$ and (1) $\sofkX{\st_2}{\klist{\kappa'}}$. Apply IH to (a) and (1).
+
+% \pfcase{(k-listrec)} $\st=\keyw{listrec}(\st_1; \st_2; \sx, \sy. \st_3)$ and (1) $\sofkX{\st_2}{\kappa}$. Apply IH to (a) and (1).
+
+% \pfcase{(k-triv)} $\kappa=\iunit$. Apply (kf-unit).
+
+% \pfcase{(k-pair)} $\st=(\st_1, \st_2)$ and $\kappa=\kappa_1\times\kappa_2$ and (1) $\sofkX{\st_1}{\kappa_1}$ and (2) $\sofkX{\st_2}{\kappa_2}$.
+By IH on (a) and (1), we have (3) $\kDelta \vdash \kappa_1$.
+By IH on (a) and (2), we have (4) $\kDelta \vdash \kappa_2$.
+Apply (kf-prod) to (3) and (4).
+
+% \pfcase{(k-fst)} $\st=\sfst{\st'}$ and (1) $\sofkX{\st'}{\kappa\times\kappa_2}$. By IH on (a) and (1), we have (2) $\kDelta \vdash \kappa \times \kappa_2$. Apply left-inversion of (kf-prod) on (2).
+
+% \pfcase{(k-snd)} $\st=\ssnd{\st'}$ and (1) $\sofkX{\st'}{\kappa_1\times\kappa}$. By IH on (a) and (1), we have (2) $\kDelta \vdash \kappa_1 \times \kappa$. Apply right-inversion of (kf-prod) on (2).
+
+% \pfcase{(k-inl)} $\st=\sinl{\kappa}{\st'}$ and (1) $\kDelta \vdash {\kappa}$. Apply (1).
+
+% \pfcase{(k-inr)} $\st=\sinr{\kappa}{\st'}$ and (1) $\kDelta \vdash {\kappa}$. Apply (1).
+
+% \pfcase{(k-raise)} $\st=\sraise{\kappa}$ and (1) $\kDelta \vdash \kappa$. Apply (1).
+
+\pfcases{(k-ty-parr) and (k-ty-ext) and (k-ty-other)} In each case, $\kappa=\kty$. Apply (kf-ty).
+
+\pfcases{(k-tycase-parr) and (k-tycase-ext)} In each case, $\st=\stycase{c}{\st'}{\sx}{\st_1}{\st_2}$ and (1) $\sofkX{\st_2}{\kappa}$. Apply IH to (a) and (1).
+
+\pfcases{[all rules in Figures \ref{fig:kinding-qity-new} and \ref{fig:kinding-qity-shared}]} In each case, $\kappa=\kity$. Apply (kf-ity).
+
+\pfcases{[all rules in Figure \ref{fig:kinding-qitm-new} and \ref{fig:kinding-qitm-shared}]} In each case, $\kappa=\kitm$. Apply (kf-itm).
+
+\pfcase{(k-ana)} We have $\kappa=\kitm$. Apply (kf-itm).
+
+\pfcase{(k-syn)} We have $\kappa=\kty\times\kitm$. Apply (kf-prod). Apply (kf-ty). Apply (kf-itm).
+\end{proof}
+
+%\subsubsection{Argument Bounds}
+\begin{lemma}[Weakening of Argument Bounds]
+\label{lemma:weakening-of-argument-bounds}
+\small If (a) $\sofkn{\kDelta}{\kGamma}{\Phi}{n'}{\st}{\kappa}$ and (b) $n' < n$ then $\sofkn{\kDelta}{\kGamma}{\Phi}{n}{\st}{\kappa}$.
+\end{lemma}
+\begin{proof}Rule induction on (a).
+
+All cases except the following proceed generically by applying the IH to all kinding premises then re-applying the same rule, because they are defined for all $n$ and all premises are checked under the same $n$.
+
+\pfcase{(k-qitm-anatrans)} In this case, $\st=\sqitm{\anatrans{n''}{\st'}}$ and (1) $n'' < n'$ and (2) $\sofkn{\kDelta}{\kGamma}{\Phi}{n'}{\st'}{\kty}$.
+By transitivity on (1) and (b), we have (3) $n'' < n$.
+By the IH on (2) and (b), we have (4) $\sofkn{\kDelta}{\kGamma}{n}{\st'}{\kty}$.
+Apply (k-qitm-anatrans) to (3) and (4).
+
+\pfcase{(k-qitm-syntrans)} In this case, $\st=\sqitm{\syntrans{n''}}$ and (1) $n'' < n'$. By transitivity on (1) and (b), we have (2) $n'' < n$. Apply (k-qitm-syntrans) to (2).
+
+\pfcase{(k-ana)} In this case, $\st=\sana{n''}{\st'}$ and (1) $n'' < n'$ and (2) $\sofkn{\kDelta}{\kGamma}{\Phi}{n'}{\st'}{\kty}$.
+By transitivity on (1) and (b), we have (3) $n'' < n$.
+By the IH on (2) and (b), we have (4) $\sofkn{\kDelta}{\kGamma}{n}{\st'}{\kty}$.
+Apply (k-ana) to (3) and (4).
+
+\pfcase{(k-syn)} In this case, $\st=\ssyn{n''}$ and (1) $n'' < n'$. By transitivity on (1) and (b), we have (2) $n'' < n$. Apply (k-syn) to (2).
+\end{proof}
+
+\subsection{Static Dynamics}
+
+Figures \ref{fig:static-dynamics-SL-core} to \ref{fig:static-dynamics-SL-qitm-anasyn} give the dynamics of the SL as a structural operational semantics.
+Although the number of rules are large, the majority of them are uninteresting.
+Figure \ref{fig:static-dynamics-SL-core} simply gives the standard dynamics of the core language (i.e.
+these rules come from \cite{pfpl}, with an $\argEnv$ tacked on to each judgement but never inspected).
+We assume standard substitution lemmas:
+
+\begin{assumption}[Static Type Variable Substitution]
+\label{assumption:static-type-variable-substitution}
+If (a) $\sofk{\kDelta, \kalpha}{\kGamma}{\Phi}{\st}{\kappa}$ and (b) ${\kDelta}\vdash {\kappa'}$ then $\sofk{\kDelta}{[\kappa'/\kalpha]\kGamma}{\Phi}{[\kappa'/\kalpha]\st}{[\kappa'/\kalpha]\kappa}$.
+\end{assumption}
+
+\begin{assumption}[Static Term Variable Substitution]
+\label{assumption:static-term-variable-substitution}
+If (a) $\sofk{\kDelta}{\kGamma, \sx :: \kappa'}{\Phi}{\st}{\kappa}$ and (b) $\sofk{\kDelta}{\kGamma}{\Phi}{\st'}{\kappa'}$ then $\sofk{\kDelta}{\kGamma}{\Phi}{[\st'/\sx]\st}{\kappa}$.
+\end{assumption}
+
+Figure \ref{fig:static-dynamics-SL-types} gives the dynamics of types and type case analysis.
+
+\begin{definition}[Types]
+\label{def:types}
+We write $\istype{\st}{\Phi}$ iff $\sofkVX{\st}{\kty}$ and $\sval{\st}{\es;\Upsilon;\Phi}$.
+\end{definition}
+
+Figures \ref{fig:static-dynamics-SL-qity-shared}, \ref{fig:static-dynamics-SL-qitm-shared-1} and \ref{fig:static-dynamics-SL-qitm-shared-2} simply recurse through forms shared between the internal language and the translational internal language, to get to one of the forms that are unique to the translational internal language, defined in Figures \ref{fig:static-dynamics-SL-qity-new} and \ref{fig:static-dynamics-SL-qitm-new}.
+The following lemmas hold:
+
+\begin{lemma}[Stepping Quoted Translational Internal Types]
+\label{lemma:stepping-quoted-translational-internal-types}
+If $\sstepA{\sqity{\qity}}{\st'}$ then $\st'=\sqity{\qity'}$.
+\end{lemma}
+\begin{proof}
+By rule induction on the assumption.
+All rules that apply are given in Figure \ref{fig:static-dynamics-SL-qity-shared}, and in all cases the right hand side of the stepping judgement is of the form $\sqity{\qity'}$.
+\end{proof}
+
+\begin{lemma}[Stepping Quoted Translational Internal Terms]
+\label{lemma:stepping-quoted-translational-internal-terms}
+If $\sstepA{\sqitm{\qitm}}{\st'}$ then $\st'=\sqitm{\qitm'}$.
+\end{lemma}
+\begin{proof}
+By rule induction on the assumption.
+All rules that apply are given in Figure \ref{fig:static-dynamics-SL-qitm-shared-1} and \ref{fig:static-dynamics-SL-qitm-shared-2}, and in all cases the right hand side of the stepping judgement is of the form $\sqitm{\qitm'}$.
+\end{proof}
+
+Figure \ref{fig:static-dynamics-SL-qitm-anasyn} gives the dynamics of the SL-EL interface.
+
+\subsubsection{Canonical Forms}
+\begin{lemma}[Static Canonical Forms]
+\label{lemma:static-canonical-forms}
+Letting $\argEnv=\es;\Upsilon;\Phi$, if (a) $\sofkn{\emptyset}{\emptyset}{\Phi}{n}{\st}{\kappa}$ and (b) $\sval{\st}{\argEnv}$ then:
+\begin{enumerate}
+\item If $\kappa=\karrow{\kappa_1}{\kappa_2}$, then (i) $\st=\slam{\kappa_1}{\sx}{\st'}$ and (ii) $\sofk{\emptyset}{\emptyset, \sx :: \kappa_1}{\Phi}{\st'}{\kappa_2}$.
+\item If $\kappa=\kforall{\kalpha}{\kappa}$, then (i) $\st=\sLam{\kalpha}{\st'}$ and (ii) $\sofk{\emptyset, \kalpha}{\emptyset}{\Phi}{\st'}{\kappa}$.
+\item If $\kappa=\klist{\kappa}$, then either (i) $\st=\keyw{nil}[\kappa]$ or (ii) (I) $\st=\keyw{cons}(\stx{hd}; \stx{tl})$ and (II) $\sofkVX{\stx{hd}}{\kappa}$ and (III) $\svalA{\stx{hd}}$ and (IV) $\sofkVX{\stx{tl}}{\klist{\kappa}}$ and (V) $\svalA{\stx{tl}}$.
+\item If $\kappa=\kunit$, then $\st=()$.
+\item If $\kappa=\kprod{\kappa_1}{\kappa_2}$ then (i) $\st=(\st_1, \st_2)$ and (ii) $\sofkVX{\st_1}{\kappa_1}$ and (iii) $\svalA{\st_1}$ and (iv) $\sofkVX{\st_2}{\kappa_2}$ and (v) $\svalA{\st_2}$. 
+\item If $\kappa=\ksum{\kappa_1}{\kappa_2}$ then either (I) (i) $\st=\sinl{\kappa}{\st'}$ and (ii) $\sofkVX{\st'}{\kappa_1}$ and (iii) $\svalA{\st'}$, or (II) (i) $\st=\sinr{\kappa}{\st'}$ and (ii) $\sofkVX{\st'}{\kappa_2}$ and (iii) $\svalA{\st'}$.
+\item If $\kappa = \kty$ then either 
+    \begin{enumerate}[(I)]
+    \item $\st = \sty{\rightharpoonup}{(\st_1, \st_2)}$ and (i) $\istype{\st_1}{\Phi}$ and (ii) $\istype{\st_2}{\Phi}$ and (iii) $\sofkVX{(\st_1, \st_2)}{\kty\times\kty}$ and (iv) $\svalA{(\st_1, \st_2)}$; or
+    \item $\st = \sty{\tc}{\sttyidx}$ and (i) $\tcdef{\tc}{\tcsig{\ktyidx}{\chi}}{\omega} \in \Phi$ and (ii) $\sofkn{\emptyset}{\emptyset}{\Phi}{0}{\st}{\ktyidx}$ and (iii) $\svalA{\sttyidx}$; or
+    \item $\st = \sty{\keyw{other}[m; \kappa]}{(\stx{data}, \sqity{\qity})}$ and (i) $\emptyset \vdash \kappa$ and (ii) $\sofkVX{\stx{data}}{\kappa}$ and (iii) $\svalA{\stx{data}}$ and (iv) $\sofkVX{\sqity{\qity}}{\kity}$ and (v) $\svalA{\sqity{\qity}}$.
+    \end{enumerate}
+\item If $\kappa = \kity$ then $\st=\sqity{\qity}$ and either 
+    \begin{enumerate}
+    \item[(I)] $\qity = \srep{\st'}$ and $\istype{\st'}{\Phi}$; or
+    \item[(II)] The outer form of $\qity$ is shared with $\tau$ and if $\qity'$ is a sub-term of $\qity$ then $\sofkn{\emptyset}{\emptyset}{\Phi}{0}{\sqity{\qity'}}{\kity}$ and $\svalNA{\sqity{\qity'}}$.
+    \end{enumerate}
+\item If $\kappa = \kitm$ then $\st=\sqitm{\qitm}$ and either 
+    \begin{enumerate}
+    \item[(I)] $\qitm=\anatrans{n'}{\st'}$ and (i) $n' < n$ and $\argEnv=\es;\Upsilon;\Phi'$ and (ii) $|\es| = n''$ and (iii) $n' < n''$ and (iv) $\istype{\st'}{\Phi}$; or
+    \item[(II)] $\qitm=\syntrans{n}$ and (i) $n' < n$ and $\argEnv=\es;\Upsilon;\Phi'$ and (ii) $|\es| = n''$ and (iii) $n' < n''$.
+    \item[(III)] The outer form of $\qitm$ is shared with $\itm$ and if $\qitm'$ is a sub-term of $\qitm$ then $\sofkn{\emptyset}{\emptyset}{\Phi}{n}{\sqitm{\qitm'}}{\kitm}$ and $\sval{\sqitm{\qitm'}}{\argEnv}$ and if $\qity$ is a sub-term of $\qitm$ then $\sofkn{\emptyset}{\emptyset}{\Phi}{0}{\sqity{\qity}}{\kity}$ and $\svalNA{\sqity{\qity}}$.
+    \end{enumerate}
+\end{enumerate}
+\end{lemma}
+\begin{proof}
+The proofs for parts 1-6 follow the usual script and are omitted \cite{pfpl}.
+
+\begin{enumerate}
+\item[7.] Rule induction on (a) and (b).
+There is one form for which a kinding rule where $\kappa$ can be $\kty$ and a value rule are both defined, $\st=\sty{c}{\sttyidx}$.
+The only value rule that applies is (n-ty-val), so (1) $\svalA{\sttyidx}$.
+Case analysis on $c$:
+
+\pfcase{$c=\rightharpoonup$} The only kinding rule that applies is (k-ty-parr), so  (2) $\sofkVX{\sttyidx}{\kty\times\kty}$.
+
+By the IH on (2) and (1), we thus have that $\sttyidx=(\st_1, \st_2)$ where (3) $\sofkVX{\st_1}{\kty}$ and (4) $\svalA{\st_1}$ and (5) $\sofkVX{\st_2}{\kty}$ and (6) $\svalA{\st_2}$.
+
+By the definition of types applied to (3) and (4), we have (7) $\istype{\st_1}{\Phi}$.
+
+By the definition of types applied to (5) and (6), we have (8) $\istype{\st_2}{\Phi}$.
+
+By (k-prod) on (3) and (5), we have (9) $\sofkVX{(\st_1, \st_2)}{\kty\times\kty}$.
+
+By (n-pair-val) on (4) and (6), we have (10) $\svalA{(\st_1, \st_2)}$.
+
+Thus, we can complete our proof by (I) with (7), (8), (9) and (10).
+
+\pfcase{$c=\tc$} The only kinding rule that applies is (k-ty-ext), so (2) $\tcdef{\tc}{\tcsig{\ktyidx}{\chi}}{\omega} \in \Phi$ and (3) $\sofkn{\emptyset}{\emptyset}{\Phi}{0}{\st}{\ktyidx}$.
+Thus, we can complete our proof by (II) with (2), (3) and (1).
+
+\pfcase{$c=\keyw{other}[m;\kappa]$} The only kinding rule that applies is (k-ty-other), so  (2) $\emptyset \vdash \kappa$ and (3) $\sofkVX{\sttyidx}{\kappa\times\kity}$.
+By the IH on (3), (b), (c) and (1), we thus have that $\sttyidx=(\stx{data}, \stx{trans})$ where (4) $\sofkVX{\stx{data}}{\kappa}$ and (5) $\svalA{\stx{data}}$ and (6) $\sofkVX{\stx{trans}}{\kity}$ and (7) $\svalA{\stx{trans}}$.
+
+By the IH on (6) and (7), we have that $\stx{trans}=\sqity{\qity}$.
+
+Thus, we can complete our proof by (III) with (2), (4), (5), (6) and (7).
+
+\item[8.] Rule induction on (a) and (b). All forms that can be given kind $\kity$ and define a value rule are of the form $\sqity{\qity}$. Case analysis on the form of $\qity$:
+
+\pfcase{$\qity=\qtuq{\st}$} There is no value rule for this form.
+
+\pfcase{$\qity=\srep{\st'}$} The only kinding rule for this form is (k-qity-trans), so  (1) $\sofkVX{\st'}{\kty}$.
+
+The only value rule for this form is (n-q-tytrans-val), so (2) $\svalA{\st'}$.
+
+By the definition of types on 1 and 2, we have (3) $\istype{\st'}{\Phi}$. We can thus complete our proof by (I) on (3).
+
+\pfcase{[$\qity$ has form shared with $\tau$]} All kinding rules for these forms are given in Figure \ref{fig:kinding-qity-shared}.
+For each form, there is one rule.
+That rule has the property that (1) if $\qity'$ is a direct sub-term of $\qity$, then there is a premise of the form $\sofkVX{\sqity{\qity'}}{\kity}$.
+
+All rules in the static dynamics for these forms are given in Figure \ref{fig:static-dynamics-SL-qity-shared}.
+For each form, there is one value rule.
+This rule has the property that (2) if $\qity'$ is a sub-term of $\qity$, the rule has a premise of the form (2) $\svalA{\sqity{\qity'}}$.
+
+We can thus complete our proof by (II) on (1) and (2) in each case.
+
+\item[9.] Rule induction on (a) and (b). All forms that can be given kind $\kitm$ and define a value rule are of the form $\sqitm{\qitm}$. Case analysis on the form of $\qitm$:
+
+\pfcase{$\qitm=\quq{\st'}$} There is no value rule for this form.
+
+\pfcase{$\qitm=\anatrans{n'}{\st'}$} The only kinding rule for this form is (k-qitm-anatrans), so (1) $n' < n$ and (2) $\sofkVX{\st'}{\kty}$.
+
+The only value rule for this form is (n-q-anatrans-val), so $\argEnv = \es;\Upsilon;\Phi'$ and (3) $\svalA{\st'}$ and (4) $|\es|=n''$ and (5) $n' < n''$.
+
+By the definition of types on (2) and (3), we have (6) $\istype{\st'}{\Phi}$.
+
+We can thus complete our proof by (I) on (1), (4), (5) and (6).
+
+\pfcase{$\qitm=\syntrans{n'}$} The only kinding rule for this form is (k-qitm-syntrans) so (1) $n' < n$.
+
+The only value rule for this form is (n-q-syntrans-val) so  $\argEnv=\es;\Upsilon;\Phi'$ and (2) $|\es|=n''$ and (3) $n' < n''$.
+
+We can thus complete our proof by (II) on (1), (2) and (3).
+
+\pfcase{[$\qitm$ has form shared with $\itm$]} All kinding rules for these forms are given in Figure \ref{fig:kinding-qitm-shared}.
+For each form, there is one rule.
+That rule has the property that (1) if $\qitm'$ is a direct sub-term of $\qitm$, then there is a premise of the form $\sofkVX{\sqitm{\qitm'}}{\kitm}$, and (2) if $\qity'$ is a direct sub-term of $\qitm$, then there is a premise of the form $\sofkVX{\sqity{\qity'}}{\kity}$.
+
+All rules in the static dynamics for these forms are given in Figure \ref{fig:static-dynamics-SL-qitm-shared-1} and \ref{fig:static-dynamics-SL-qitm-shared-2}.
+For each form, there is one value rule.
+This rule has the property that (3) if $\qitm'$ is a sub-term of $\qitm$, then there is a premise of the form $\svalA{\sqitm{\qitm'}}$, and (4) if $\qity'$ is a sub-term of $\qitm$, then there is a premise of the form $\svalA{\sqity{\qity'}}$.
+
+We can thus complete our proof by (III) on (1), (2), (3) and (4) in each case.
+\end{enumerate}
+\end{proof}
+
+%\subsubsection{Argument Independence}
+As suggested by Lemma \ref{lemma:static-canonical-forms}, only static values of kind $\kitm$ are argument-dependent (i.e.
+are indexed by a natural number).
+It is helpful to have a lemma that says that values of other kinds, in particular equality kinds, are argument independent.
+
+\begin{lemma}[Values of Equality Kind are Argument Independent]
+\label{lemma:values-of-equality-kind-are-argument-independent}
+If (a) $\sofkVX{\st}{\kappa}$ and (b) $\svalA{\st}$ and (c) $\keq{}{\kappa}$ then (i) $\sofkz{\emptyset}{\emptyset}{\Phi}{\st}{\kappa}$ and (ii) $\sval{\st}{\argEnv'}$.
+\end{lemma}
+\begin{proof}
+By rule induction on (c).
+
+\pfcase{(keq-list)} $\kappa=\klist{\kappa'}$ and $\keq{}{\kappa'}$.
+Rule induction on (a) and (b).
+There are only twos forms that can have kind $\klist{\kappa'}$ and define value rules, $\st=\keyw{nil}[\kappa']$ and $\st=\keyw{cons}(\st_1; \st_2)$.
+The only kinding rules that apply are (k-nil) and (k-cons), respectively.
+The only value rules that apply are (n-nil-val) and (n-cons-val), respectively.
+All of these rules are defined for all $n$ and all $\argEnv$, and only kind check against equality kinds $\kappa'$ and $\klist{\kappa'}$ in their premises, so we can simply re-apply them inductively to conclude (i) and (ii) in each case.
+
+\pfcase{(keq-unit)} $\kappa=\kunit$.
+Rule induction on (a) and (b).
+There is only one form that can have kind $\kunit$ and defines a value rule, $\st=()$.
+The only kinding rule that applies is (k-unit) and the only value rule that applies is (n-triv-val).
+Both are defined for all $n$ and all $\argEnv$, and have no premises, so we can simply re-apply them to conclude (i) and (ii).
+
+\pfcase{(keq-prod)} $\kappa=\kappa_1\times\kappa_2$ and $\keq{}{\kappa_1}$ and $\keq{}{\kappa_2}$.
+Rule induction on (a) and (b).
+There is only one form that can have kind $\kappa_1\times\kappa_2$ and defines a value rule, $\st=(\st_1, \st_2)$.
+The only kinding rule that applies is (k-pair).
+The only value rule that applies is (n-pair-val).
+Both are defined for all $n$ and all $\argEnv$, and only kind check against equality kinds $\kappa_1$ or $\kappa_2$ in the premises, so we can simply re-apply them inductively to conclude (i) and (ii).
+
+\pfcase{(keq-sum)} $\kappa=\kappa_1+\kappa_2$.
+Rule induction on (a) and (b).
+There are two forms that can have kind $\kappa_1 + \kappa_2$ and define value rules, $\st=\sinl{\kappa}{\st_1}$ and $\st=\sinr{\kappa}{\st_2}$.
+The only kinding rules that apply are (k-inl) and (k-inr), respectively.
+The only value rules that apply are (n-inl-val) and (n-inr-val), respectively.
+All of these rules are defined for all $n$ and all $\argEnv$, and only kind check against equality kinds $\kappa_1$ or $\kappa_2$ in the premises, so we can simply re-apply them inductively to conclude (i) and (ii) in each case.
+
+\pfcase{(keq-ty)} $\kappa=\kty$.
+Rule induction on (a) and (b).
+There is one form that can have kind $\kty$ and defines value rules, $\st=\sty{c}{\sttyidx}$.
+The only value rule that applies is (n-ty-val), so we have (1) $\svalA{\sttyidx}$.
+Three kinding rules might apply:
+\begin{quote}
+\pfcase{(k-ty-parr)} In this case, $c=\rightharpoonup$ and (2) $\sofkVX{\sttyidx}{\kty\times\kty}$.
+By (keq-prod) and (keq-ty) and (keq-ty), we have that (3) $\keq{}{\kty\times\kty}$.
+Applying the IH to (2), (1) and (3), we have that (4) $\sofkz{\emptyset}{\emptyset}{\Phi}{\sttyidx}{\kty\times\kty}$ and (6) $\sval{\sttyidx}{\argEnv'}$.
+By (k-ty-parr) on (5), we have (i).
+By (n-ty-val) on (6), we have (ii).
+
+\pfcase{(k-ty-ext)} In this case, $c=\tc$ and (2) $\tcdef{\tc}{\tcsig{\ktyidx}{\chi}}{\theta} \in\Phi$ and (3) $\sofkVX{\sttyidx}{\ktyidx}$.
+By Lemma \ref{lemma:tycons-have-valid-signatures} on (2), we have that (4) $\vdash \tcsig{\ktyidx}{\chi}$.
+By Lemma \ref{lemma:sigs-define-eq-type-index-kinds} on (4), we have that (5) $\keq{}{\ktyidx}$.
+By the IH on (3), (1) and (5) we have (6) $\sofkz{\emptyset}{\emptyset}{\Phi}{\sttyidx}{\ktyidx}$ and (7) $\sval{\sttyidx}{\argEnv'}$.
+Applying (k-ty-ext) to (2) and (6), we have (i).
+Applying (n-ty-val) to (1), we have (ii).
+
+\pfcase{(k-ty-other)} In this case, $c=\keyw{other}[m;\kappa']$ and (2) $\keq{}{\kappa'}$ and (3) $\sofkVX{\sttyidx}{\kappa'\times\kity}$.
+By (keq-ity), we have (4) $\keq{}{\kity}$.
+By (keq-prod) on (2) and (4), we have (5) $\keq{}{\kappa'\times\kity}$.
+By the IH on (3), (1) and (5), we have (6) $\sofkz{\emptyset}{\emptyset}{\sttyidx}{\kappa'\times\kity}$ and (7) $\sval{\sttyidx}{\argEnv'}$.
+By (k-ty-other) on (2) and (6), we have (i).
+By (n-ty-val) on (7), we have (ii).
+\end{quote}
+
+\pfcase{(keq-ity)} $\kappa=\kity$.
+Rule induction on (a) and (b).
+The only forms that can have kind $\kity$ and define value rules are of the form $\sqity{\qity}$.
+The kinding rules for these are defined in Figures \ref{fig:kinding-qity-new} and \ref{fig:kinding-qity-shared}.
+\begin{quote}
+\pfcase{[k-qity-trans]} In this case, $\qity=\srep{\st'}$ and (1) $\sofkVX{\st'}{\kty}$.
+The only value rule that applies is (n-q-tytrans-val), so (2) $\svalA{\st'}$.
+Note that (3) $\keq{}{\kty}$ by (keq-ty).
+By the IH on (1), (2) and (3), we have (4) $\sofkz{\emptyset}{\emptyset}{\Phi}{\st'}{\kty}$ and (5) $\sval{\st'}{\argEnv'}$.
+By (k-qity-trans) on (4), we have (i).
+By (n-q-tytrans-val) on (5), we have (ii).
+
+\pfcase{[all rules in Figure \ref{fig:kinding-qity-shared}]} Observe that all rules where $\qity$ is of shared form are defined for all $n$, and that for all sub-terms $\qity'$ of $\qity$, there is a premise of the form $\sofkVX{\sqity{\qity'}}{\kity}$, and there are no other premises.
+
+The value rules for these are defined in Figures \ref{fig:static-dynamics-SL-qity-shared}.
+Observe that all rules are defined for all $\argEnv$ and for all sub-terms $\qity'$ of $\qity$, there is a premise of the form $\svalA{\sqity{\qity'}}$, and there are no other premises.
+
+Thus, in every case, we can apply the IH to each sub-term of $\qity$, then reapply the same rules to derive (i) and (ii), respectively.
+\end{quote}
+\end{proof}
+
+\begin{lemma}[Types are Argument Independent]
+\label{lemma:types-are-argument-independent}
+$\istype{\st}{\Phi}$ iff (a) $\sofkz{\emptyset}{\emptyset}{\Phi}{\st}{\kty}$ and (b) $\sval{\st}{\cdot;\emptyset;\Phi}$.
+\end{lemma}
+\begin{proof}
+In the forward direction, by the definition of $\istype{\st}{\Phi}$ we have (1) $\sofkVX{\st}{\kty}$ and (2) $\sval{\st}{\es;\Upsilon;\Phi}$.
+By (keq-ty), we have (3) $\keq{}{\kty}$.
+By Lemma \ref{lemma:values-of-equality-kind-are-argument-independent} on 1, 2 and 3, we have (a) and (b).
+
+In the backwards direction, apply the definition of $\istype{\st}{\Phi}$ to (a) and (b).
+\end{proof}
+
+\subsection{Kind Safety}
+Kind safety ensures that normalization of well-kinded static terms cannot go wrong. We can take a standard progress and preservation based approach.
+
+\begin{theorem}[Static Progress]\label{thm:static-progress}
+Letting $\argEnv=\es;\Upsilon;\Phi$, if (a) $\sofkn{\emptyset}{\emptyset}{\Phi}{n}{\st}{\kappa}$ and (b) $|\es|=n$  then (I) $\sstep{\st}{\argEnv}{\st'}$ or (II) $\sval{\st}{\argEnv}$ or (III) $\serr{\st}{\argEnv}$.
 \end{theorem}
-\begin{lemma}[Internal Preservation]
-If $\emptyset~\emptyset \vdash \iota : \tau$ and $\iota \mapsto \iota'$ then $\emptyset~\emptyset \vdash \iota' : \tau$. 
-\end{lemma}
-
-We assume substitution and simultaneous $n$-ary substitutions, $\delta$ and $\gamma$, have the standard semantics, and that typing and type formation contexts, $\Delta$ and $\Gamma$, obey standard properties of weakening, exchange and contraction.
-We assume that the names of variables and type variables are unimportant, so that $\alpha$-equivalent terms, types and contexts can be identified implicitly throughout this work.
-In particular, we need the definitions of substitution validity shown in Figures \ref{fig:internal-type-substitution-validity} and \ref{fig:internal-term-substitution-validity} and the following lemmas.
-
-\begin{figure}[t]
-\small
-\hspace{-5px}\input{syntax-table-IL}
-\caption{Syntax of {$\mathcal{L}\{{\rightharpoonup}\,{\forall}\,{\mu}\,{\iunit}\,{\times}\,{+}\}$}, our internal language (IL).
-Metavariable $x$ ranges over term variables and $\alpha$ and $t$ both range over type variables.}
-\label{syntax-IL}
-\end{figure}
-
-\begin{figure}
-\small\fbox{$\Delta \vdash \delta : \Delta$}~~$\delta ::=  \emptyset ~|~ \delta, \tau/\alpha$
-
-\begin{mathpar}
-\inferrule[tysub-emp]{ }{\Delta \vdash \emptyset : \emptyset}
-
-\inferrule[tysub-ext]{
-    \Delta \vdash \delta : \Delta'\\
-    \Delta \vdash \tau
-}{
-    \Delta \vdash \delta, \tau/\alpha : \Delta', \alpha
-}
-\end{mathpar}
-\caption{Internal Type Substitution Validity}
-\label{fig:internal-type-substitution-validity}
-\end{figure}
-\begin{figure}
-\small\fbox{$\Delta~\Gamma \vdash \gamma : \Gamma$}~~$\gamma ::= \emptyset ~|~ \gamma, \iota/x$
-\begin{mathpar}
-\inferrule[tmsub-emp]{ }{\Delta~\Gamma \vdash \emptyset : \emptyset}
-
-\inferrule[tmsub-ext]{
-    \Delta~\Gamma \vdash \gamma : \Gamma'\\
-    \Delta \vdash \tau\\
-    \Delta~\Gamma \vdash \iota : \tau
-}{
-    \Delta~\Gamma \vdash \gamma, \iota/x : \Gamma', x : \tau
-}
-\end{mathpar}
-\caption{Internal Term Substitution Validity}
-\label{fig:internal-term-substitution-validity}
-\end{figure}
-\begin{lemma}[Internal Type Substitution on Types] If $\Delta \vdash \delta : \Delta'$ and $\Delta\Delta' \vdash \tau$ then $\Delta \vdash [\delta]\tau$.\end{lemma}
-
-\begin{lemma}[Internal Type Substitutions on Typing Contexts] If $\Delta \vdash \delta : \Delta'$ and $\Delta\Delta' \vdash \Gamma$ then $\Delta \vdash [\delta]\Gamma$.
-\end{lemma}
-
-\begin{lemma}[Internal Type Substitutions on Terms] If $\Delta \vdash \delta : \Delta'$ and $\Delta\Delta' \vdash \tau$ and $\Delta\Delta' \vdash \Gamma$ and $\Delta\Delta'~\Gamma \vdash \iota : \tau$ then $\Delta~[\delta]\Gamma \vdash [\delta]\iota : [\delta]\tau$.
-\end{lemma}
-
-\begin{lemma}[Internal Term Substitutions] If $\Delta \vdash \Gamma$ and $\Delta \vdash \Gamma'$ and $\Delta~\Gamma \vdash \gamma : \Gamma'$ and $\Delta~\Gamma\Gamma' \vdash \iota : \tau$ then $\Delta~\Gamma \vdash [\gamma]\iota : \tau$.
-\end{lemma}
-
-\section{Tycon Contexts}
-\begin{figure}\fbox{$\vdash \Phi$}\vspace{-25px}
-\begin{mathpar}
-\small
-\inferrule[tcc-emp]{ }{\vdash \cdot}
-
-\inferrule[tcc-ext]{
-    \vdash \Phi\\
-    \tc \notin \text{dom}(\Phi)\\
-    \keq{\emptyset}{\ktyidx}\\
-    \sofkz{\emptyset}{\emptyset}{\Phi}{\stx{schema}}{\ktyidx \rightarrow \kity}\\\\
-    \vdash_{\Phi,  \tcdef{\tc}{\tcsig{\ktyidx}{\chi}}{\tcstruct{\stx{schema}}{\omega}}} \omega \sim \tcsig{\ktyidx}{\chi}
-}{
-    \vdash \Phi, \tcdef{\tc}{\tcsig{\ktyidx}{\chi}}{\tcstruct{\stx{schema}}{\omega}}
-}
-\end{mathpar}
-\caption{Tycon context well-definedness.}
-\label{fig:tycon-ctxs}
-\end{figure}
-\begin{figure}[t]
-\small\fbox{$\vdash_\Phi \omega \sim \psi$}
-\begin{mathpar}\small
-\inferrule[ocstruct-intro]{
-    \introsig{\klitidx} \in \chi\\
-    \emptyset \vdash \klitidx\\\\
-    \sofkz{\emptyset}{\emptyset}{\Phi}{\stx{def}}{\ktyidx \rightarrow \klitidx \rightarrow \kargs \rightarrow \kitm}
-}{
-    \vdash_\Phi \tcstructn{\stx{def}} \sim \tcsig{\ktyidx}{\chi}
-}
-
-\inferrule[ocstruct-targ]{
-    \vdash_\Phi \omega \sim \tcsig{\ktyidx}{\chi}\\
-    \opname{op}\notin\text{dom}(\chi)\\
-    \emptyset \vdash \klitidx\\\\
-    \sofkn{\emptyset}{\emptyset}{\Phi}{0}{\stx{def}}{\ktyidx \rightarrow \klitidx \rightarrow \kargs \rightarrow (\kty \times \kitm)}
-}{
-    \vdash_\Phi \tcstructc{\omega}{op}{\stx{def}} \sim \tcsig{\ktyidx}{\chi, \opsig{\opname{op}}{\klitidx}}
-}
-\end{mathpar}
-\caption{Checking opcon structures against tycon signatures.}
-\label{ocstruct}
-\end{figure}
-
-\subsection{Tycon Context Well-Definedness}
-
-\subsection{Equality Kinds}
-Need an equational theory for SL to state equality kind property, but not important for other metatheory.
-
-\subsection{Full Examples}
-
-\section{Static Language}
-\subsection{Kind Formation}
-
-\subsection{Kinding Context Formation}
-
-\subsection{Kinding}
-
-\subsection{Dynamic Semantics}
-
-\subsection{Kind Safety}
-
-\section{Types}
-\subsection{Type Translations}
-
-\subsection{Typing Context Translations}
-
-\paragraph{Unicity}
-The rules are structured so that if a term is well-typed, both its type and translation are unique.\todo{could move this whole thing to supplement if room needed}
-\begin{theorem}[Unicity]
-If $\vdash \Phi$ and $\vdash_\Phi \Upsilon \leadsto \Gamma$ and $\vdash_\Phi \st \leadsto \tau$ and $\vdash_\Phi \st' \leadsto \tau'$ and $\eanaX{e}{\st}{\iota}$ and $\eanaX{e}{\st'}{\iota'}$ then $\st = \st'$ and $\tau=\tau'$ and $\iota = \iota'$.
+\begin{proof} Rule induction on (a).
+
+\pfcases{[All rules in Figure \ref{fig:kinding-core}]} The proof of progress for core language constructs follows the standard script \cite{pfpl} and is omitted.
+
+\pfcase{(k-ty-parr)} In this case, we have $\st=\sty{\rightharpoonup}{\sttyidx}$ and $\kappa=\kty$ and (1) $\sofkVX{\sttyidx}{\kty\times\kty}$. Applying the IH to (1) and (b), we have three cases:
+\begin{quote}
+    \pfcase{(2) $\sstepA{\sttyidx}{\sttyidx'}$} Applying (n-ty-step) to (2), we have (3) $\sstepA{\sty{\rightharpoonup}{\sttyidx}}{\sty{\rightharpoonup}{\sttyidx'}}$. We can conclude by (I) on (3).
+
+    \pfcase{(2) $\svalA{\sttyidx}$} Applying (n-ty-val) to (2), we have (3) $\svalA{\sty{\rightharpoonup}{\sttyidx}}$. We can conclude by (II) on (3).
+
+    \pfcase{(2) $\serrA{\sttyidx}$} Applying (n-ty-err-prop) to (2), we have (3) $\serrA{\sty{\rightharpoonup}{\sttyidx}}$. We can conclude by (III) on (3).
+\end{quote}
+
+\pfcase{(k-ty-ext)} In this case, we have $\st=\sty{\tc}{\sttyidx}$ and $\kappa=\kty$  and (1) $\sofkVX{\sttyidx}{\ktyidx}$. Applying the IH to (1) and (b), we have three cases:
+\begin{quote}
+    \pfcase{(2) $\sstepA{\sttyidx}{\sttyidx'}$} Applying (n-ty-step) to (2), we have (3) $\sstepA{\sty{\tc}{\sttyidx}}{\sty{\rightharpoonup}{\sttyidx'}}$. We can conclude by (I) on (3).
+
+    \pfcase{(2) $\svalA{\sttyidx}$} Applying (n-ty-val) to (2), we have (3) $\svalA{\sty{\tc}{\sttyidx}}$. We can conclude by (II) on (3).
+
+    \pfcase{(2) $\serrA{\sttyidx}$} Applying (n-ty-err-prop) to (2), we have (3) $\serrA{\sty{\tc}{\sttyidx}}$. We can conclude by (III) on (3).
+\end{quote}
+
+\pfcase{(k-ty-other)} In this case, we have $\st=\sty{\otherc{m}{\kappa'}}{\sttyidx}$ and $\kappa=\kty$ and (1) $\sofkVX{\sttyidx}{\kappa'\times\kity}$.
+Applying the IH to (1) and (b), we have three cases:
+\begin{quote}
+    \pfcase{(2) $\sstepA{\sttyidx}{\sttyidx'}$} Applying (n-ty-step) to (2), we have (3) $\sstepA{\sty{\tc}{\sttyidx}}{\sty{\otherc{m}{\kappa'}}{\sttyidx'}}$. We can conclude by (I) on (3).
+
+    \pfcase{(2) $\svalA{\sttyidx}$} Applying (n-ty-val) to (2), we have (3) $\svalA{\sty{\otherc{m}{\kappa'}}{\sttyidx}}$. We can conclude by (II) on (3).
+
+    \pfcase{(2) $\serrA{\sttyidx}$} Applying (n-ty-err-prop) to (2), we have (3) $\serrA{\sty{\otherc{m}{\kappa'}}{\sttyidx}}$. We can conclude by (III) on (3).
+\end{quote}
+
+\pfcase{(k-tycase-parr)} In this case, we have $\st=\stycase{\rightharpoonup}{\st'}{\sx}{\st_1}{\st_2}$ and (1) $\sofkVX{\st'}{\kty}$. Applying the IH to (1) and (b), we have three cases:
+\begin{quote}
+    \pfcase{(2) $\sstepA{\st'}{\st''}$} Applying (n-tycase-step) to (2), we have (3) $$\sstepA{\stycase{{\rightharpoonup}}{\st'}{\sx}{\st_1}{\st_2}}{\stycase{\rightharpoonup}{\st''}{\sx}{\st_1}{\st_2}}$$ We can conclude by (I) on (3).
+
+    \pfcase{(2) $\svalA{\st'}$} By Lemma \ref{lemma:static-canonical-forms}.7 on (1) and (2), we have three cases: 
+    \begin{quote}
+        \pfcase{$\st'=\sty{\rightharpoonup}{(\st_1', \st_2')}$} Applying (n-tycase-elim-1) to (2), we have (3) $$\sstepA{\stycase{\rightharpoonup}{\sty{\rightharpoonup}{(\st_1', \st_2')}}{\sx}{\st_1}{\st_2}}{[(\st_1', \st_2')/\sx]\st_1}$$ We can conclude by (I) on (3).
+
+        \pfcase{$\st'=\sty{\tc}{\sttyidx}$} We have (3) ${\rightharpoonup}\neq\tc$.
+Applying (n-tycase-elim-2) to (2) and (3), we have (4) $$\sstepA{\stycase{\rightharpoonup}{\sty{\tc}{\sttyidx}}{\sx}{\st_1}{\st_2}}{\st_2}$$ We can conclude by (I) on (4).
+
+        \pfcase{$\st'=\sty{\otherc{m}{\kappa'}}{(\stx{data}, \sqity{\qity})}$} We have  (3) ${\rightharpoonup}\neq\otherc{m}{\kappa'}$.
+Applying (n-tycase-elim-2) to (2) and (3), we have (4) $$\sstepA{\stycase{\rightharpoonup}{\sty{\otherc{m}{\kappa'}}{(\stx{data}, \sqity{\qity})}}{\sx}{\st_1}{\st_2}}{\st_2}$$ We can conclude by (I) on (4).
+    \end{quote}
+
+    \pfcase{(2) $\serrA{\st'}$} Applying (n-tycase-err-prop) to (2), we have (3) $$\serrA{\stycase{\rightharpoonup}{\st'}{\sx}{\st_1}{\st_2}}$$ We can conclude by (III) on (3). 
+\end{quote}
+
+\pfcase{(k-tycase-ext)} In this case, we have $\st=\stycase{\tc}{\st'}{\sx}{\st_1}{\st_2}$ and (1) $\sofkVX{\st'}{\kty}$. Applying the IH to (1) and (b), we have  three cases:
+\begin{quote}
+    \pfcase{(2) $\sstepA{\st'}{\st''}$} Applying (n-tycase-step) to (2), we have (3) $$\sstepA{\stycase{{\tc}}{\st'}{\sx}{\st_1}{\st_2}}{\stycase{\tc}{\st''}{\sx}{\st_1}{\st_2}}$$ We can conclude by (I) on (3).
+
+    \pfcase{(2) $\svalA{\st'}$} By Lemma \ref{lemma:static-canonical-forms}.7 on (1) and (2), we have three cases: 
+    \begin{quote}
+        \pfcase{$\st'=\sty{\rightharpoonup}{(\st_1', \st_2')}$} We have (3) $\tc\neq{\rightharpoonup}$.
+Applying (n-tycase-elim-2) to (2) and (3), we have (4) $$\sstepA{\stycase{\tc}{\sty{\rightharpoonup}{(\st_1', \st_2')}}{\sx}{\st_1}{\st_2}}{\st_2}$$ We can conclude by (I) on (4).
+
+        \pfcase{$\st'=\sty{\tc'}{\sttyidx}$ and (3) $\svalA{\sttyidx}$} There are two cases:
+        \begin{quote}
+            \pfcase{$\tc=\tc'$} Applying (n-tycase-elim-1) to (3), we have (4) $$\sstepA{\stycase{\tc}{\sty{\tc}{\sttyidx}}{\sx}{\st_1}{\st_2}}{[\sttyidx/\sx]\st_1}$$ We can conclude by (I) on (4).
+
+            \pfcase{(4) $\tc\neq\tc'$} Applying (n-tycase-elim-2) to (3) and (4), we have (5) $$\sstepA{\stycase{\tc}{\sty{\tc'}{\sttyidx}}{\sx}{\st_1}{\st_2}}{\st_2}$$ We can conclude by (I) on (5).
+        \end{quote}
+
+        \pfcase{$\st'=\sty{\otherc{m}{\kappa'}}{(\stx{data}, \sqity{\qity})}$} We have  (3) $\tc\neq\otherc{m}{\kappa'}$.
+Applying (n-tycase-elim-2) to (2) and (3), we have (4) $$\sstepA{\stycase{\rightharpoonup}{\sty{\otherc{m}{\kappa'}}{(\stx{data}, \sqity{\qity})}}{\sx}{\st_1}{\st_2}}{\st_2}$$ We can conclude by (I) on (4).
+    \end{quote}
+
+    \pfcase{(2) $\serrA{\st'}$} Applying (n-tycase-err-prop) to (2), we have (3) $$\serrA{\stycase{\rightharpoonup}{\st'}{\sx}{\st_1}{\st_2}}$$ We can conclude by (III) on (3). 
+\end{quote}
+
+\pfcase{(k-qity-unquote)} We have $\st=\sqity{\qtuq{\st'}}$ and (1) $\sofkVX{\st'}{\kity}$. Applying the IH to (1) and (b), we have three cases:
+
+\begin{quote}
+    \pfcase{(2) $\sstepA{\st'}{\st''}$} Applying (n-q-t-unquote-step) to (2), we have (3) $\sstepA{\sqity{\qtuq{\st'}}}{\sqity{\qtuq{\st''}}}$. We can conclude by (I) on (3).
+
+    \pfcase{(2) $\svalA{\st'}$} Applying Lemma \ref{lemma:static-canonical-forms}.8 to (1) and (2), we have that (3) $\st'=\sqity{\qity}$.
+Applying (n-q-t-unquote-elim) to (3), we have (4) $\sstepA{\sqity{\qtuq{\sqity{\qity}}}}{\sqity{\qity}}$.
+We can conclude by (I) on (4).
+
+    \pfcase{(2) $\serrA{\st'}$} Applying (n-q-t-unquote-err-prop) on (2), we have (3) $\serrA{\sqity{\qtuq{\st'}}}$. We can conclude by (III) on (3).
+\end{quote}
+
+\pfcase{(k-qity-trans)} We have $\st=\sqity{\srep{\st'}}$ and (1) $\sofkVX{\st'}{\kty}$. Applying the IH to (1) and (b), we have three cases:
+
+\begin{quote}
+    \pfcase{(2) $\sstepA{\st'}{\st''}$} Applying (n-q-tytrans-step) to (2), we have (3) $\sstepA{\sqity{\srep{\st'}}}{\sqity{\srep{\st''}}}$. We can conclude by (I) on (3).
+
+    \pfcase{(2) $\svalA{\st'}$} Applying (n-q-tytrans-val) to (2), we have (3) $\svalA{\sqity{\srep{\st'}}}$. We can conclude by (I) on (3).
+
+    \pfcase{(2) $\serrA{\st'}$} Applying (n-q-tytrans-err-prop) on (2), we have (3) $\serrA{\sqity{\qtuq{\st'}}}$. We can conclude by (III) on (3).
+\end{quote}
+
+\pfcases{[All rules in Figure \ref{fig:kinding-qity-shared}]} We give an example base case and inductive cases with 0, 1 and 2 sub-terms.
+All rules with the same number of sub-terms follow the corresponding script, replacing only the forms and rule names:
+
+\begin{quote}
+    \pfcase{(k-qity-alpha)} We have that $\st=\sqity{\alpha}$. Applying (n-q-alpha-val), we have (1) $\svalA{\sqity{\alpha}}$. We can conclude by (II) on (1).
+
+    \pfcase{(k-qity-forall)} We have that $\st=\sqity{\iforall{\alpha}{\qity}}$ and (1) $\sofkVX{\sqity{\qity}}{\kity}$. Applying the IH to (1) and (b), we have three cases:
+
+    \begin{quote}
+        \pfcase{(2) $\sstepA{\sqity{\qity}}{\st'}$} By Lemma \ref{lemma:stepping-quoted-translational-internal-types} on (2), we have $\st'=\sqity{\qity'}$.
+Applying (n-q-forall-step) to (2), we have (3) $\sstepA{\sqity{\iforall{\alpha}{\qity}}}{\sqity{\iforall{\alpha}{\qity'}}}$.
+We can conclude by (I) on (3).
+
+        \pfcase{(2) $\svalA{\sqity{\qity}}$} Applying (n-q-forall-val) to (2), we have (3) $\svalA{\sqity{\iforall{\alpha}{\qity}}}$. We can conclude by (II) on (3).
+
+        \pfcase{(2) $\serrA{\sqity{\qity}}$} Applying (n-q-forall-err-prop) to (2), we have (3) $\serrA{\sqity{\iforall{\alpha}{\qity}}}$. We can conclude by (III) on (3).
+    \end{quote}
+
+    \pfcase{(k-qity-parr)} We have that $\st=\sqity{\iparr{\qity_1}{\qity_2}}$ and (1) $\sofkVX{\sqity{\qity_1}}{\kity}$ and (2) $\sofkVX{\sqity{\qity_2}}{\kity}$.
+Applying the IH to (1), there are three possibilities:
+
+    \begin{quote}
+        \pfcase{(3) $\sstepA{\sqity{\qity_1}}{\st_1'}$} By Lemma \ref{lemma:stepping-quoted-translational-internal-types} on (3), $\st_1'=\sqity{\qity_1'}$.
+Applying (n-q-parr-step-1) to (3), we have (4) $\sstepA{\sqity{\iparr{\qity_1}{\qity_2}}}{\sqity{\iparr{\qity_1'}{\qity_2}}}$.
+We can conclude by (I) on (4).
+
+        \pfcase{(3) $\svalA{\sqity{\qity_1}}$} Applying the IH to (2), there are three possibilities:
+
+        \begin{quote}
+            \pfcase{(4) $\sstepA{\sqity{\qity_2}}{\st_2'}$} By Lemma \ref{lemma:stepping-quoted-translational-internal-types} on (4), $\st_2'=\sqity{\qity_2'}$.
+By (n-q-parr-step-2) on (3) and (4), we have (5) $\sstepA{\sqity{\iparr{\qity_1}{\qity_2}}}{\sqity{\iparr{\qity_1}{\qity_2'}}}$.
+We can conclude by (I) on (5).
+
+            \pfcase{(4) $\svalA{\sqity{\qity_2}}$} Applying (n-q-parr-val) to (3) and (4), we have (5) $\svalA{\sqity{\iparr{\qity_1}{\qity_2}}}$. We can conclude by (II) on (5).
+
+            \pfcase{(4) $\serrA{\sqity{\qity_2}}$} Applying (n-q-parr-err-prop-2) to (4), we have (5) $\serrA{\sqity{\iparr{\qity_1}{\qity_2}}}$. We can conclude by (III) on (5).
+        \end{quote}
+
+        \pfcase{(3) $\serrA{\sqity{\qity_1}}$} Applying (n-q-parr-err-prop-1) to (3), we have (4) $\serrA{\sqity{\iparr{\qity_1}{\qity_2}}}$. We can conclude by (III) on (4).
+    \end{quote}
+\end{quote}
+
+% \pfcase{(k-qity-t)} We have that $\st=\sqity{t}$. Applying (n-q-t-val), we have (1) $\svalA{\sqity{t}}$. We can conclude by (II) on (1).
+
+% \pfcase{(k-qity-mu)} We have that $\st=\sqity{\imu{t}{\qity}}$ and (1) $\sofkVX{\sqity{\qity}}{\kity}$. Applying the IH to (1) and (b), we have three cases:
+
+% \begin{quote}
+%     \pfcase{(2) $\sstepA{\sqity{\qity}}{\st'}$} By Lemma \ref{lemma:stepping-quoted-translational-internal-types} on (2), we have $\st'=\sqity{\qity'}$.
+Applying (n-q-mu-step) to (2), we have (3) $\sstepA{\sqity{\imu{t}{\qity}}}{\sqity{\imu{t}{\qity'}}}$.
+We can conclude by (I) on (3).
+
+%     \pfcase{(2) $\svalA{\sqity{\qity}}$} Applying (n-q-mu-val) to (2), we have (3) $\svalA{\sqity{\imu{t}{\qity}}}$. We can conclude by (II) on (3).
+
+%     \pfcase{(2) $\serrA{\sqity{\qity}}$} Applying (n-q-mu-err-prop) to (2), we have (3) $\serrA{\sqity{\imu{t}{\qity}}}$. We can conclude by (III) on (3).
+% \end{quote}
+
+% \pfcase{(k-qity-unit)} We have that $\st=\sqity{()}$. Applying (n-q-unit-val), we have (1) $\svalA{\sqity{())}}$. We can conclude by (II) on (1).
+
+% \pfcase{(k-qity-prod)} We have that $\st=\sqity{\iprod{\qity_1}{\qity_2}}$ and (1) $\sofkVX{\sqity{\qity_1}}{\kity}$ and (2) $\sofkVX{\sqity{\qity_2}}{\kity}$.
+Applying the IH to (1), there are three possibilities:
+
+% \begin{quote}
+%     \pfcase{(3) $\sstepA{\sqity{\qity_1}}{\st_1'}$} By Lemma \ref{lemma:stepping-quoted-translational-internal-types} on (3), $\st_1'=\sqity{\qity_1'}$.
+Applying (n-q-prod-step-1) to (3), we have (4) $\sstepA{\sqity{\iprod{\qity_1}{\qity_2}}}{\sqity{\iprod{\qity_1'}{\qity_2}}}$.
+We can conclude by (I) on (4).
+
+%     \pfcase{(3) $\svalA{\sqity{\qity_1}}$} Applying the IH to (2), there are three possibilities:
+
+%     \begin{quote}
+%         \pfcase{(4) $\sstepA{\sqity{\qity_2}}{\st_2'}$} By Lemma \ref{lemma:stepping-quoted-translational-internal-types} on (4), $\st_2'=\sqity{\qity_2'}$.
+By (n-q-prod-step-2) on (3) and (4), we have (5) $\sstepA{\sqity{\iprod{\qity_1}{\qity_2}}}{\sqity{\iprod{\qity_1}{\qity_2'}}}$.
+We can conclude by (I) on (5).
+
+%         \pfcase{(4) $\svalA{\sqity{\qity_2}}$} Applying (n-q-prod-val) to (3) and (4), we have (5) $\svalA{\sqity{\iprod{\qity_1}{\qity_2}}}$. We can conclude by (II) on (5).
+
+%         \pfcase{(4) $\serrA{\sqity{\qity_2}}$} Applying (n-q-prod-err-prop-2) to (4), we have (5) $\serrA{\sqity{\iprod{\qity_1}{\qity_2}}}$. We can conclude by (III) on (5).
+%     \end{quote}
+
+%     \pfcase{(3) $\serrA{\sqity{\qity_1}}$} Applying (n-q-prod-err-prop-1) to (3), we have (4) $\serrA{\sqity{\iprod{\qity_1}{\qity_2}}}$. We can conclude by (III) on (4).
+% \end{quote}
+
+% \pfcase{(k-qity-sum)} We have that $\st=\sqity{\isum{\qity_1}{\qity_2}}$ and (1) $\sofkVX{\sqity{\qity_1}}{\kity}$ and (2) $\sofkVX{\sqity{\qity_2}}{\kity}$.
+Applying the IH to (1), there are three possibilities:
+
+% \begin{quote}
+%     \pfcase{(3) $\sstepA{\sqity{\qity_1}}{\st_1'}$} By Lemma \ref{lemma:stepping-quoted-translational-internal-types} on (3), $\st_1'=\sqity{\qity_1'}$.
+Applying (n-q-sum-step-1) to (3), we have (4) $\sstepA{\sqity{\isum{\qity_1}{\qity_2}}}{\sqity{\isum{\qity_1'}{\qity_2}}}$.
+We can conclude by (I) on (4).
+
+%     \pfcase{(3) $\svalA{\sqity{\qity_1}}$} Applying the IH to (2), there are three possibilities:
+
+%     \begin{quote}
+%         \pfcase{(4) $\sstepA{\sqity{\qity_2}}{\st_2'}$} By Lemma \ref{lemma:stepping-quoted-translational-internal-types} on (4), $\st_2'=\sqity{\qity_2'}$.
+By (n-q-sum-step-2) on (3) and (4), we have (5) $\sstepA{\sqity{\isum{\qity_1}{\qity_2}}}{\sqity{\isum{\qity_1}{\qity_2'}}}$.
+We can conclude by (I) on (5).
+
+%         \pfcase{(4) $\svalA{\sqity{\qity_2}}$} Applying (n-q-sum-val) to (3) and (4), we have (5) $\svalA{\sqity{\isum{\qity_1}{\qity_2}}}$. We can conclude by (II) on (5).
+
+%         \pfcase{(4) $\serrA{\sqity{\qity_2}}$} Applying (n-q-sum-err-prop-2) to (4), we have (5) $\serrA{\sqity{\isum{\qity_1}{\qity_2}}}$. We can conclude by (III) on (5).
+%     \end{quote}
+
+%     \pfcase{(3) $\serrA{\sqity{\qity_1}}$} Applying (n-q-sum-err-prop-1) to (3), we have (4) $\serrA{\sqity{\isum{\qity_1}{\qity_2}}}$. We can conclude by (III) on (4).
+% \end{quote}
+
+\pfcase{(k-qitm-unquote)} We have $\st=\sqitm{\quq{\st'}}$ and (1) $\sofkVX{\st'}{\kitm}$. Applying the IH to (1) and (b), we have three cases:
+
+\begin{quote}
+    \pfcase{(2) $\sstepA{\st'}{\st''}$} Applying (n-q-unquote-step) to (2), we have (3) $\sstepA{\sqitm{\quq{\st'}}}{\sqitm{\quq{\st''}}}$. We can conclude by (I) on (3).
+
+    \pfcase{(2) $\svalA{\st'}$} Applying Lemma \ref{lemma:static-canonical-forms}.9 to (1) and (2), we have that (3) $\st'=\sqitm{\qitm}$.
+Applying (n-q-unquote-elim) to (3), we have (4) $\sstepA{\sqitm{\quq{\sqitm{\qitm}}}}{\sqitm{\qitm}}$.
+We can conclude by (I) on (4).
+
+    \pfcase{(2) $\serrA{\st'}$} Applying (n-q-unquote-err-prop) on (2), we have (3) $\serrA{\sqitm{\quq{\st'}}}$. We can conclude by (III) on (3).
+\end{quote}
+
+\pfcase{(k-qitm-anatrans)} We have $\st=\sqitm{\anatrans{n'}{\st'}}$ and (1) $n' < n$ and (2) $\sofkVX{\st'}{\kty}$. Applying the IH to (2), we have three cases:
+
+\begin{quote}
+    \pfcase{(3) $\sstepA{\st'}{\st''}$} Applying (n-q-anatrans-step) to (3), we have (4) $\sstepA{\sqitm{\anatrans{n'}{\st'}}}{\sqitm{\anatrans{n'}{\st''}}}$. We can conclude by (I) on (4).
+
+    \pfcase{(3) $\svalA{\st'}$} Applying (n-q-anatrans-val) to (2), (b) and (1), we have (4) $\svalA{\sqitm{\anatrans{n'}{\st'}}}$. We can conclude by (II) on (4).
+
+    \pfcase{(3) $\serrA{\st'}$} Applying (n-q-anatrans-err-prop) to (3), we have (4) $\serrA{\sqitm{\anatrans{n'}{\st'}}}$. We can conclude by (III) on (4).
+\end{quote}
+
+\pfcase{(k-qitm-syntrans)} We have $\st=\sqitm{\syntrans{n'}}$ and (1) $n' < n$.
+Applying (n-q-syntrans-val) to (b) and (1), we have (2) $\svalA{\sqitm{\syntrans{n'}}}$.
+We can conclude by (II) on (2).
+
+\pfcases{[All rules in Figure \ref{fig:kinding-qitm-shared}]} We give an example base case and inductive cases with 0, 1, 2 and 3 sub-terms.
+All rules with the same number of sub-terms follow the corresponding script, replacing only the forms and rule names:
+
+\begin{quote}
+    \pfcase{[k-qitm-var]} In this case, $\st=\sqitm{x}$. Applying (n-q-x-val), we have (1) $\svalA{\sqitm{x}}$. We can conclude by (II) on (1).
+
+    \pfcase{(k-qitm-tabs)} We have that $\st=\sqitm{\iLam{\alpha}{\qitm}}$ and (1) $\sofkVX{\sqitm{\qitm}}{\kitm}$. Applying the IH to (1) and (b), we have three cases:
+
+    \begin{quote}
+        \pfcase{(2) $\sstepA{\sqitm{\qitm}}{\st'}$} By Lemma \ref{lemma:stepping-quoted-translational-internal-terms} on (2), we have $\st'=\sqitm{\qitm'}$.
+Applying (n-q-tabs-step) to (2), we have (3) $\sstepA{\sqitm{\iLam{\alpha}{\qitm}}}{\sqitm{\iLam{\alpha}{\qitm'}}}$.
+We can conclude by (I) on (3).
+
+        \pfcase{(2) $\svalA{\sqitm{\qitm}}$} Applying (n-q-tabs-val) to (2), we have (3) $\svalA{\sqitm{\iLam{\alpha}{\qitm}}}$. We can conclude by (II) on (3).
+
+        \pfcase{(2) $\serrA{\sqitm{\qitm}}$} Applying (n-q-tabs-err-prop) to (2), we have (3) $\serrA{\sqitm{\iLam{\alpha}{\qitm}}}$. We can conclude by (III) on (3).
+    \end{quote}
+
+    \pfcase{[k-qitm-abs]} In this case, $\st=\sqitm{\ilam{\qity}{x}{\qitm}}$ and (1) $\sofkVX{\sqity{\qity}}{\kity}$ and (2) $\sofkVX{\sqitm{\qitm}}{\kitm}$.
+Applying the IH to (1) and (b), we have three cases:
+
+    \begin{quote}
+        \pfcase{(3) $\sstepA{\sqity{\qity}}{\st_1'}$} By Lemma \ref{lemma:stepping-quoted-translational-internal-types} on (3), $\st_1'=\sqity{\qity'}$.
+Applying (n-q-abs-step-1) on (3), we have (4) $\sstepA{\sqitm{\ilam{\qity}{x}{\qitm}}}{\sqitm{\ilam{\qity'}{x}{\qitm}}}$.
+We can conclude by (I) on (4).
+
+        \pfcase{(3) $\svalA{\sqity{\qity}}$} Applying the IH (2) and (b), we have three cases:
+        \begin{quote}
+            \pfcase{(4) $\sstepA{\sqitm{\qitm}}{\st_2'}$} By Lemma \ref{lemma:stepping-quoted-translational-internal-terms} on (4), we have $\st_2'=\sqitm{\qitm'}$.
+Applying (n-q-abs-step-2) to (3) and (4), we have (5) $\qstep{\ilam{\qity}{x}{\qitm}}{\ilam{\qity}{x}{\qitm'}}$.
+We can conclude by (I) on (5).
+
+            \pfcase{(4) $\qval{\qitm}$} Applying (n-q-abs-val) to (3) and (4), we have (5) $\qval{\ilam{\qity}{x}{\qitm}}$. We can conclude by (II) on (5).
+
+            \pfcase{(4) $\qerr{\qitm}$} Applying (n-q-abs-err-prop-2) to (4), we have (5) $\qerr{\ilam{\qity}{x}{\qitm}}$. We can conclude by (III) on (5).
+        \end{quote}
+
+        \pfcase{(3) $\serrA{\sqity{\qity}}$} Applying (n-q-abs-err-prop-2) to (3), we have (4) $\qerr{\ilam{\qity}{x}{\qitm}}$. We can conclude by (III) on (4).
+    \end{quote}
+
+    \pfcase{(k-qitm-case)} In this case $\st=\sqitm{\icase{\qitm}{x}{\qitm_1}{x}{\qitm_2}}$ and (1) $\sofkVX{\sqitm{\qitm}}{\kitm}$ and (2) $\sofkVX{\sqitm{\qitm_1}}{\kitm}$ and (3) $\sofkVX{\sqitm{\qitm_2}}{\kitm}$.
+Applying the IH to (1) and (b), we have three cases:
+
+    \begin{quote}
+        \pfcase{(4) $\sstepA{\sqitm{\qitm}}{\st'}$} By Lemma \ref{lemma:stepping-quoted-translational-internal-terms} on (4), $\st'=\sqitm{\qitm'}$.
+Applying (n-q-case-step-1) to (4), we have (5) $\sstepA{\sqitm{\icase{\qitm}{x}{\qitm_1}{x}{\qitm_2}}}{\sqitm{\icase{\qitm'}{x}{\qitm_1}{x}{\qitm_2}}}$.
+We can conclude by (I) on (5).
+
+        \pfcase{(4) $\svalA{\sqitm{\qitm}}$} Applying the IH to (2) and (b), we have three cases:
+
+        \begin{quote}
+            \pfcase{(5) $\sstepA{\sqitm{\qitm_1}}{\st_1'}$} By Lemma \ref{lemma:stepping-quoted-translational-internal-terms} on (5), $\st_1'=\sqitm{\qitm_1'}$.
+Applying (n-q-case-step-2) to (4) and (5), we have (6) $\sstepA{\sqitm{\icase{\qitm}{x}{\qitm_1}{x}{\qitm_2}}}{\sqitm{\icase{\qitm}{x}{\qitm_1'}{x}{\qitm_2}}}$.
+We can conclude by (I) on (6).
+
+            \pfcase{(5) $\svalA{\sqitm{\qitm_1}}$} Applying the IH to (3) and (b), we have three cases:
+
+            \begin{quote}
+                \pfcase{(6) $\sstepA{\sqitm{\qitm_2}}{\st_2'}$} By Lemma \ref{lemma:stepping-quoted-translational-internal-terms} on (6), $\st_2'=\sqitm{\qitm_2'}$.
+Applying (n-q-case-step-3) to (4), (5) and (6), we have (7) $\sstepA{\sqitm{\icase{\qitm}{x}{\qitm_1}{x}{\qitm_2}}}{\sqitm{\icase{\qitm}{x}{\qitm_1}{x}{\qitm_2'}}}$.
+We can conclude by (I) on (7).
+
+                \pfcase{(6) $\svalA{\sqitm{\qitm_2}}$} Applying (n-q-case-val) to (4), (5) and (6), we have (7) $\svalA{\sqitm{\icase{\qitm}{x}{\qitm_1}{x}{\qitm_2}}}$. We can conclude by (II) on (7).
+
+                \pfcase{(6) $\serrA{\sqitm{\qitm_2}}$} Applyig (n-q-case-err-prop-3) to (6), we have (7) $\serrA{\sqitm{\icase{\qitm}{x}{\qitm_1}{x}{\qitm_2}}}$. We can conclude by (III) on (7).
+            \end{quote}
+
+            \pfcase{(5) $\serrA{\sqitm{\qitm_1}}$} Applying (n-q-case-err-prop-2) to (5), we have (6) $\serrA{\sqitm{\icase{\qitm}{x}{\qitm_1}{x}{\qitm_2}}}$. We can conclude by (III) on (6).
+        \end{quote}
+
+        \pfcase{(4) $\serrA{\sqitm{\qitm}}$} Applying (n-q-case-err-prop-1) to (4), we have (5) $\serrA{\sqitm{\icase{\qitm}{x}{\qitm_1}{x}{\qitm_2}}}$. We can conclude by (III) on (5).
+    \end{quote}
+
+\end{quote}
+
+% \pfcase{[k-qitm-ap]} In this case, $\st=\sqitm{\iap{\st_1}{\st_2}}$ and (1) $\sofkVX{\sqitm{\qitm_1}}$ and (2) $\sofkVX{\sqitm{\qitm_2}}$. Applying the IH to (1) and (b), we have three cases:
+
+% \begin{quote}
+%     \pfcase{(3) $\sstepA{\sqitm{\qitm_1}}{\st_1'}$} By Lemma \ref{lemma:stepping-quoted-translational-internal-terms} on (3), $\st_1'=\sqitm{\qitm_1'}$.
+Applying (n-q-ap-step-1) on (3), we have (4) $\sstepA{\sqitm{\iap{\qitm_1}{\qitm_2}}}{\sqitm{\iap{\qitm_1'}{\qitm_2}}}$.
+We can conclude by (I) on (4).
+
+%     \pfcase{(3) $\svalA{\sqitm{\qitm_1}}$} Applying the IH (2) and (b), we have three cases:
+%     \begin{quote}
+%         \pfcase{(4) $\sstepA{\sqitm{\qitm_2}}{\st_2'}$} By Lemma \ref{lemma:stepping-quoted-translational-internal-terms} on (4), we have $\st_2'=\sqitm{\qitm_2'}$.
+Applying (n-q-ap-step-2) to (3) and (4), we have (5) $\qstep{\iap{\qitm_1}{\qitm_2}}{\iap{\qitm_1}{\qitm_2'}}$.
+We can conclude by (I) on (5).
+
+%         \pfcase{(4) $\qval{\qitm_2}$} Applying (n-q-ap-val) to (3) and (4), we have (5) $\qval{\iap{\qitm_1}{\qitm_2}}$. We can conclude by (II) on (5).
+
+%         \pfcase{(4) $\qerr{\qitm_2}$} Applying (n-q-ap-err-prop-2) to (4), we have (5) $\qerr{\iap{\qitm_1}{\qitm_2}}$. We can conclude by (III) on (5).
+%     \end{quote}
+
+%     \pfcase{(3) $\serrA{\sqitm{\qitm_1}}$} Applying (n-q-ap-err-prop-2) to (3), we have (4) $\qerr{\iap{\qitm_1}{\qitm_2}}$. We can conclude by (III) on (4).
+% \end{quote}
+
+% \pfcase{(k-qitm-tap)} We have that $\st=\sqitm{\iAp{\qity}{\qitm}}$ and (1) $\sofkVX{\sqitm{\qitm}}{\kitm}$ and (2) $\sofkVX{\sqity{\qity}}{\kity}$.
+Applying the IH to (1), there are three possibilities:
+
+% \begin{quote}
+%     \pfcase{(3) $\sstepA{\sqitm{\qitm}}{\st_1'}$} By Lemma \ref{lemma:stepping-quoted-translational-internal-terms} on (3), $\st_1'=\sqitm{\qitm'}$.
+Applying (n-q-tap-step-1) to (3), we have (4) $\sstepA{\sqitm{\iAp{\qity}{\qitm}}}{\sqitm{\iAp{\qity}{\qitm'}}}$.
+We can conclude by (I) on (4).
+
+%     \pfcase{(3) $\svalA{\sqitm{\qitm}}$} Applying the IH to (2), there are three possibilities:
+
+%     \begin{quote}
+%         \pfcase{(4) $\sstepA{\sqity{\qity}}{\st_2'}$} By Lemma \ref{lemma:stepping-quoted-translational-internal-types} on (4), $\st_2'=\sqity{\qity'}$.
+By (n-q-tap-step-2) on (3) and (4), we have (5) $\sstepA{\sqity{\iAp{\qity}{\qitm}}}{\sqity{\iAp{\qity'}{\qitm}}}$.
+We can conclude by (I) on (5).
+
+%         \pfcase{(4) $\svalA{\sqity{\qity}}$} Applying (n-q-tap-val) to (3) and (4), we have (5) $\svalA{\sqity{\iAp{\qity}{\qitm}}}$. We can conclude by (II) on (5).
+
+%         \pfcase{(4) $\serrA{\sqity{\qity}}$} Applying (n-q-tap-err-prop-2) to (4), we have (5) $\serrA{\sqity{\iAp{\qity}{\qitm}}}$. We can conclude by (III) on (5).
+%     \end{quote}
+
+%     \pfcase{(3) $\serrA{\sqitm{\qitm}}$} Applying (n-q-tap-err-prop-1) to (3), we have (4) $\serrA{\sqitm{\iAp{\qity}{\qitm}}}$. We can conclude by (III) on (4).
+% \end{quote}
+
+% \pfcase{[k-qitm-fold]} In this case, $\st=\sqitm{\ifold{t}{\qity}{\qitm}}$ and (1) $\sofkVX{\sqity{\qity}}{\kity}$ and (2) $\sofkVX{\sqitm{\qitm}}{\kitm}$.
+Applying the IH to (1) and (b), we have three cases:
+
+% \begin{quote}
+%     \pfcase{(3) $\sstepA{\sqity{\qity}}{\st_1'}$} By Lemma \ref{lemma:stepping-quoted-translational-internal-types} on (3), $\st_1'=\sqity{\qity'}$.
+Applying (n-q-fold-step-1) on (3), we have (4) $\sstepA{\sqitm{\ifold{t}{\qity}{\qitm}}}{\sqitm{\ifold{t}{\qity'}{\qitm}}}$.
+We can conclude by (I) on (4).
+
+%     \pfcase{(3) $\svalA{\sqity{\qity}}$} Applying the IH (2) and (b), we have three cases:
+%     \begin{quote}
+%         \pfcase{(4) $\sstepA{\sqitm{\qitm}}{\st_2'}$} By Lemma \ref{lemma:stepping-quoted-translational-internal-terms} on (4), we have $\st_2'=\sqitm{\qitm'}$.
+Applying (n-q-fold-step-2) to (3) and (4), we have (5) $\qstep{\ifold{t}{\qity}{\qitm}}{\ifold{t}{\qity}{\qitm'}}$.
+We can conclude by (I) on (5).
+
+%         \pfcase{(4) $\qval{\qitm}$} Applying (n-q-fold-val) to (3) and (4), we have (5) $\qval{\ifold{t}{\qity}{\qitm}}$. We can conclude by (II) on (5).
+
+%         \pfcase{(4) $\qerr{\qitm}$} Applying (n-q-fold-err-prop-2) to (4), we have (5) $\qerr{\ifold{t}{\qity}{\qitm}}$. We can conclude by (III) on (5).
+%     \end{quote}
+
+%     \pfcase{(3) $\serrA{\sqity{\qity}}$} Applying (n-q-fold-err-prop-2) to (3), we have (4) $\qerr{\ifold{t}{\qity}{\qitm}}$. We can conclude by (III) on (4).
+% \end{quote}
+
+% \pfcase{(k-qitm-unfold)} We have that $\st=\sqitm{\iunfold{\qitm}}$ and (1) $\sofkVX{\sqitm{\qitm}}{\kitm}$. Applying the IH to (1) and (b), we have three cases:
+
+% \begin{quote}
+%     \pfcase{(2) $\sstepA{\sqitm{\qitm}}{\st'}$} By Lemma \ref{lemma:stepping-quoted-translational-internal-terms} on (2), we have $\st'=\sqitm{\qitm'}$.
+Applying (n-q-unfold-step) to (2), we have (3) $\sstepA{\sqitm{\iunfold{\qitm}}}{\sqitm{\iunfold{\qitm'}}}$.
+We can conclude by (I) on (3).
+
+%     \pfcase{(2) $\svalA{\sqitm{\qitm}}$} Applying (n-q-unfold-val) to (2), we have (3) $\svalA{\sqitm{\iunfold{\qitm}}}$. We can conclude by (II) on (3).
+
+%     \pfcase{(2) $\serrA{\sqitm{\qitm}}$} Applying (n-q-unfold-err-prop) to (2), we have (3) $\serrA{\sqitm{\iunfold{\qitm}}}$. We can conclude by (III) on (3).
+% \end{quote}
+
+% \pfcase{[k-qitm-triv]} In this case, $\st=\sqitm{())}$. Applying (n-q-triv-val), we have (1) $\svalA{\sqitm{()}}$. We can conclude by (II) on (1).
+
+% \pfcase{[k-qitm-pair]} In this case, $\st=\sqitm{\ipair{\st_1}{\st_2}}$ and (1) $\sofkVX{\sqitm{\qitm_1}}$ and (2) $\sofkVX{\sqitm{\qitm_2}}$. Applying the IH to (1) and (b), we have three cases:
+
+% \begin{quote}
+%     \pfcase{(3) $\sstepA{\sqitm{\qitm_1}}{\st_1'}$} By Lemma \ref{lemma:stepping-quoted-translational-internal-terms} on (3), $\st_1'=\sqitm{\qitm_1'}$.
+Applying (n-q-pair-step-1) on (3), we have (4) $\sstepA{\sqitm{\ipair{\qitm_1}{\qitm_2}}}{\sqitm{\ipair{\qitm_1'}{\qitm_2}}}$.
+We can conclude by (I) on (4).
+
+%     \pfcase{(3) $\svalA{\sqitm{\qitm_1}}$} Applying the IH (2) and (b), we have three cases:
+%     \begin{quote}
+%         \pfcase{(4) $\sstepA{\sqitm{\qitm_2}}{\st_2'}$} By Lemma \ref{lemma:stepping-quoted-translational-internal-terms} on (4), we have $\st_2'=\sqitm{\qitm_2'}$.
+Applying (n-q-pair-step-2) to (3) and (4), we have (5) $\qstep{\ipair{\qitm_1}{\qitm_2}}{\ipair{\qitm_1}{\qitm_2'}}$.
+We can conclude by (I) on (5).
+
+%         \pfcase{(4) $\qval{\qitm_2}$} Applying (n-q-pair-val) to (3) and (4), we have (5) $\qval{\ipair{\qitm_1}{\qitm_2}}$. We can conclude by (II) on (5).
+
+%         \pfcase{(4) $\qerr{\qitm_2}$} Applying (n-q-pair-err-prop-2) to (4), we have (5) $\qerr{\ipair{\qitm_1}{\qitm_2}}$. We can conclude by (III) on (5).
+%     \end{quote}
+
+%     \pfcase{(3) $\serrA{\sqitm{\qitm_1}}$} Applying (n-q-pair-err-prop-2) to (3), we have (4) $\qerr{\ipair{\qitm_1}{\qitm_2}}$. We can conclude by (III) on (4).
+% \end{quote}
+
+% \pfcase{(k-qitm-fst)} We have that $\st=\sqitm{\ifst{\qitm}}$ and (1) $\sofkVX{\sqitm{\qitm}}{\kitm}$. Applying the IH to (1) and (b), we have three cases:
+
+% \begin{quote}
+%     \pfcase{(2) $\sstepA{\sqitm{\qitm}}{\st'}$} By Lemma \ref{lemma:stepping-quoted-translational-internal-terms} on (2), we have $\st'=\sqitm{\qitm'}$.
+Applying (n-q-fst-step) to (2), we have (3) $\sstepA{\sqitm{\ifst{\qitm}}}{\sqitm{\ifst{\qitm'}}}$.
+We can conclude by (I) on (3).
+
+%     \pfcase{(2) $\svalA{\sqitm{\qitm}}$} Applying (n-q-fst-val) to (2), we have (3) $\svalA{\sqitm{\ifst{\qitm}}}$. We can conclude by (II) on (3).
+
+%     \pfcase{(2) $\serrA{\sqitm{\qitm}}$} Applying (n-q-fst-err-prop) to (2), we have (3) $\serrA{\sqitm{\ifst{\qitm}}}$. We can conclude by (III) on (3).
+% \end{quote}
+
+% \pfcase{(k-qitm-snd)} We have that $\st=\sqitm{\isnd{\qitm}}$ and (1) $\sofkVX{\sqitm{\qitm}}{\kitm}$. Applying the IH to (1) and (b), we have three cases:
+
+% \begin{quote}
+%     \pfcase{(2) $\sstepA{\sqitm{\qitm}}{\st'}$} By Lemma \ref{lemma:stepping-quoted-translational-internal-terms} on (2), we have $\st'=\sqitm{\qitm'}$.
+Applying (n-q-snd-step) to (2), we have (3) $\sstepA{\sqitm{\isnd{\qitm}}}{\sqitm{\isnd{\qitm'}}}$.
+We can conclude by (I) on (3).
+
+%     \pfcase{(2) $\svalA{\sqitm{\qitm}}$} Applying (n-q-snd-val) to (2), we have (3) $\svalA{\sqitm{\isnd{\qitm}}}$. We can conclude by (II) on (3).
+
+%     \pfcase{(2) $\serrA{\sqitm{\qitm}}$} Applying (n-q-snd-err-prop) to (2), we have (3) $\serrA{\sqitm{\isnd{\qitm}}}$. We can conclude by (III) on (3).
+% \end{quote}
+
+% \pfcase{[k-qitm-inl]} In this case, $\st=\sqitm{\iinl{\qity}{\qitm}}$ and (1) $\sofkVX{\sqity{\qity}}{\kity}$ and (2) $\sofkVX{\sqitm{\qitm}}{\kitm}$.
+Applying the IH to (1) and (b), we have three cases:
+
+% \begin{quote}
+%     \pfcase{(3) $\sstepA{\sqity{\qity}}{\st_1'}$} By Lemma \ref{lemma:stepping-quoted-translational-internal-types} on (3), $\st_1'=\sqity{\qity'}$.
+Applying (n-q-inl-step-1) on (3), we have (4) $\sstepA{\sqitm{\iinl{\qity}{\qitm}}}{\sqitm{\iinl{\qity'}{\qitm}}}$.
+We can conclude by (I) on (4).
+
+%     \pfcase{(3) $\svalA{\sqity{\qity}}$} Applying the IH (2) and (b), we have three cases:
+%     \begin{quote}
+%         \pfcase{(4) $\sstepA{\sqitm{\qitm}}{\st_2'}$} By Lemma \ref{lemma:stepping-quoted-translational-internal-terms} on (4), we have $\st_2'=\sqitm{\qitm'}$.
+Applying (n-q-inl-step-2) to (3) and (4), we have (5) $\qstep{\iinl{\qity}{\qitm}}{\iinl{\qity}{\qitm'}}$.
+We can conclude by (I) on (5).
+
+%         \pfcase{(4) $\qval{\qitm}$} Applying (n-q-inl-val) to (3) and (4), we have (5) $\qval{\iinl{\qity}{\qitm}}$. We can conclude by (II) on (5).
+
+%         \pfcase{(4) $\qerr{\qitm}$} Applying (n-q-inl-err-prop-2) to (4), we have (5) $\qerr{\iinl{\qity}{\qitm}}$. We can conclude by (III) on (5).
+%     \end{quote}
+
+%     \pfcase{(3) $\serrA{\sqity{\qity}}$} Applying (n-q-inl-err-prop-2) to (3), we have (4) $\qerr{\iinl{\qity}{\qitm}}$. We can conclude by (III) on (4).
+% \end{quote}
+
+% \pfcase{[k-qitm-inr]} In this case, $\st=\sqitm{\iinr{\qity}{\qitm}}$ and (1) $\sofkVX{\sqity{\qity}}{\kity}$ and (2) $\sofkVX{\sqitm{\qitm}}{\kitm}$.
+Applying the IH to (1) and (b), we have three cases:
+
+% \begin{quote}
+%     \pfcase{(3) $\sstepA{\sqity{\qity}}{\st_1'}$} By Lemma \ref{lemma:stepping-quoted-translational-internal-types} on (3), $\st_1'=\sqity{\qity'}$.
+Applying (n-q-inr-step-1) on (3), we have (4) $\sstepA{\sqitm{\iinr{\qity}{\qitm}}}{\sqitm{\iinr{\qity'}{\qitm}}}$.
+We can conclude by (I) on (4).
+
+%     \pfcase{(3) $\svalA{\sqity{\qity}}$} Applying the IH (2) and (b), we have three cases:
+%     \begin{quote}
+%         \pfcase{(4) $\sstepA{\sqitm{\qitm}}{\st_2'}$} By Lemma \ref{lemma:stepping-quoted-translational-internal-terms} on (4), we have $\st_2'=\sqitm{\qitm'}$.
+Applying (n-q-inr-step-2) to (3) and (4), we have (5) $\qstep{\iinr{\qity}{\qitm}}{\iinr{\qity}{\qitm'}}$.
+We can conclude by (I) on (5).
+
+%         \pfcase{(4) $\qval{\qitm}$} Applying (n-q-inr-val) to (3) and (4), we have (5) $\qval{\iinr{\qity}{\qitm}}$. We can conclude by (II) on (5).
+
+%         \pfcase{(4) $\qerr{\qitm}$} Applying (n-q-inr-err-prop-2) to (4), we have (5) $\qerr{\iinr{\qity}{\qitm}}$. We can conclude by (III) on (5).
+%     \end{quote}
+
+%     \pfcase{(3) $\serrA{\sqity{\qity}}$} Applying (n-q-inr-err-prop-2) to (3), we have (4) $\qerr{\iinr{\qity}{\qitm}}$. We can conclude by (III) on (4).
+% \end{quote}
+
+\pfcase{(k-ana)} We have $\st=\sana{n'}{\st'}$ and (1) $n' < n$ and (2) $\sofkVX{\st'}{\kty}$. Applying the IH to (2) and (b), we have three cases:
+\begin{quote}
+    \pfcase{(3) $\sstepA{\st'}{\st''}$} Applying (n-ana-step) to (3), we have (4) $\sstepA{\sana{n'}{\st'}}{\sana{n'}{\st''}}$. We can conclude by (I) on (4).
+
+    \pfcase{(3) $\svalA{\st'}$} By (1) and (b), we have (4) $\keyw{nth}[n'](\es)=e$.
+
+    Because our semantics are non-deterministic (the bracketed premise of (n-ana-fail) is assumed to be only advisory to implementors), by applying (n-ana-fail) to (3) and (4), we have (5) $\serrA{\sana{n'}{\st'}}$.
+We can conclude by (III) on (5).
+
+    \pfcase{(3) $\serrA{\st'}$} Applying (n-ana-err-prop) to (3), we have (4) $\serrA{\sana{n'}{\st'}}$. We can conclude by (III) on (4).
+\end{quote}
+
+\pfcase{(k-syn)} We have $\st=\ssyn{n'}$ and (1) $n' < n$.
+By (1) and (b), we have (2) $\keyw{nth}[n](\es)=e$.
+Because our semantics are non-deterministic (the bracketed premise of (n-syn-fail) is assumed to be only advisory to implementors), by applying (n-syn-fail) to (2), we have (3) $\serrA{\ssyn{n'}}$.
+We can conclude by (III) on (3).
+\end{proof}
+
+\begin{theorem}[Static Preservation]\label{thm:static-preservation}
+Letting $\argEnv=\es;\Upsilon;\Phi$, if (a) $\sstep{\st}{\es; \Upsilon; \Phi}{\st'}$ and (b) $\sofkn{\emptyset}{\emptyset}{\Phi}{n}{\st}{\kappa}$ and (c) $\vdash \Phi$ and (d) $\Upsilon~\mathtt{ctx}_{\Phi}$ then $\sofkn{\emptyset}{\emptyset}{\Phi}{n}{\st'}{\kappa}$.
 \end{theorem}
-
+\begin{proof} Rule induction on (a).
+
+\pfcase{[All rules in Figure \ref{fig:static-dynamics-SL-core}]} The cases for the core language follow the standard script and are omitted for concision.
+
+\pfcase{(n-ty-step)} In this case, we have $\st=\sty{c}{\sttyidx}$ and $\st'=\sty{c}{\sttyidx'}$ and $\kappa=\kty$ and (1) $\sstepA{\sttyidx}{\sttyidx'}$.
+Rule induction on (b).
+Only three rules apply:
+\begin{quote}
+    \pfcase{(k-ty-parr)} In this case, we have $c={\rightharpoonup}$ and (2) $\sofkVX{\sttyidx}{\kty\times\kty}$.
+Applying the IH to (1), (2), (c) and (d), we have (3) $\sofkVX{\sttyidx'}{\kty\times\kty}$.
+Apply (k-ty-parr) to (3).
+
+    \pfcase{(k-ty-ext)} In this case, we have $c=\tc$ and (2) $\tcdef{\tc}{\tcsig{\ktyidx}{\chi}}{\theta} \in \Phi$ and (3) $\sofkVX{\sttyidx}{\ktyidx}$.
+Applying the IH to (1), (3), (c) and (d), we have (4) $\sofkVX{\sttyidx'}{\ktyidx}$.
+Apply (k-ty-ext) to (2) and (4).
+
+    \pfcase{(k-ty-other)} In this case, we have $c=\otherc{m}{\kappa'}$ and (2) $\keq{}{\kappa'}$ and (3) $\sofkVX{\sttyidx}{\kappa'\times\kity}$.
+Applying the IH to (1) and (3), we have (4) $\sofkVX{\sttyidx'}{\kappa'\times\kity}$.
+Apply (k-ty-other) to (2) and (4).
+\end{quote}
+
+\pfcase{(n-tycase-step)} In this case, we have $\st=\stycase{c}{\st''}{\sx}{\st_1}{\sx}{\st_2}$ and $\st'=\stycase{c}{\st'''}{\sx}{\st_1}{\sx}{\st_2}$ and (1) $\sstepA{\st''}{\st'''}$.
+Rule induction on (b).
+Only two rules apply:
+\begin{quote}
+    \pfcase{(k-tycase-parr)} In this case, we have $c={\rightharpoonup}$ and (2) $\sofkVX{\st''}{\kty}$ and (3) $\sofk{\emptyset}{\emptyset, \sx :: \kty\times\kty}{\Phi}{\st_1}{\kappa}$ and (4) $\sofk{\emptyset}{\emptyset}{\Phi}{\st_2}{\kappa}$.
+Applying the IH to (1) and (2), we have that (5) $\sofkVX{\st'''}{\kty}$.
+Apply (k-tycase-parr) to (5), (3) and (4).
+
+    \pfcase{(k-tycase-ext)} In this case, have $c=\tc$ and (2) $\sofkVX{\st''}{\kty}$ and (3) $\tcdef{\tc}{\tcsig{\ktyidx}{\chi}}{\theta} \in \Phi$ and (4) $\sofk{\emptyset}{\emptyset, \sx :: \ktyidx}{\Phi}{\st_1}{\kappa}$ and (5) $\sofkVX{\st_2}{\kappa}$.
+Applying the IH to (1) and (2), we have that (6) $\sofkVX{\st'''}{\kty}$.
+Apply (k-tycase-ext) to (6), (3), (4) and (5).
+\end{quote}
+
+\pfcase{(n-tycase-elim-1)} In this case, we have $\st=\stycase{c}{\sty{c}{\sttyidx}}{\sx}{\st_1}{\st_2}$ and $\st'=[\sttyidx/\sx]\st_1$ and (1) $\svalA{\sty{c}{\sttyidx}}$.
+Rule induction on (b).
+Only two rules apply:
+\begin{quote}
+    \pfcase{(k-tycase-parr)} In this case, we have $c={\rightharpoonup}$ and (2) $\sofkVX{\sty{\rightharpoonup}{\sttyidx}}{\kty}$ and (3) $\sofk{\emptyset}{\emptyset, \sx :: \kty\times\kty}{\Phi}{\st_1}{\kappa}$.
+Applying Lemma \ref{lemma:static-canonical-forms}.7 to (1) and (2), we have that (4) $\sttyidx=(\st_1, \st_2)$ and (5) $\sofkVX{\sttyidx}{\kty\times\kty}$.
+Apply Assumption \ref{assumption:static-term-variable-substitution} to (3) and (5).
+
+    \pfcase{(k-tycase-ext)} In this case, we have $c=\tc$ and (2) $\sofkVX{\sty{\tc}{\sttyidx}}{\kty}$ and (3) $\tcdef{\tc}{\tcsig{\ktyidx}{\chi}}{\theta}$ and (4) $\sofk{\emptyset}{\emptyset, \sx :: \ktyidx}{\st_1}{\kappa}$.
+Applying Lemma 18.7 to (1) and (2), we have that (5) $\sofkVX{\sttyidx}{\ktyidx}$.
+Apply Assumption \ref{assumption:static-term-variable-substitution} to (4) and (5).
+\end{quote}
+
+\pfcase{(n-tycase-elim-2)} In this case, we have $\st=\stycase{c}{\sty{c'}{\sttyidx}}{\sx}{\st_1}{\st_2}$ and $\st'=\st_2$. Rule induction on (b). Only two rules apply:
+\begin{quote}
+    \pfcase{(k-tycase-parr)} In this case, we have (1) $\sofkVX{\st_2}{\kappa}$. We can conclude by (1).
+
+    \pfcase{(k-tycase-ext)} In this case, we have (1) $\sofkVX{\st_2}{\kappa}$. We can conclude by (1).
+\end{quote}
+
+\pfcase{(n-q-t-unquote-step)} In this case, we have $\st=\sqity{\qtuq{\st''}}$ and $\st'=\sqity{\qtuq{\st'''}}$ and $\kappa=\kity$ and (1) $\sstepA{\st''}{\st'''}$.
+Rule induction on (b).
+Only rule (k-qity-unquote) applies, so (2) $\sofkVX{\st''}{\kity}$.
+Applying the IH to (1) and (2), we have (3) $\sofkVX{\st'''}{\kity}$.
+Apply (k-qity-unquote) to (3).
+
+\pfcase{(n-q-t-unquote-elim)} In this case, we have $\st=\sqity{\qtuq{\sqity{\qity}}}$ and $\st'=\sqity{\qity}$ and $\kappa=\kity$.
+Rule induction on (b).
+Only rule (k-qity-unquote) applies, so (1) $\sofkVX{\sqity{\qity}}{\kity}$.
+We can conclude by (1).
+
+\pfcase{(n-q-tytrans-step)} In this case, we have $\st=\sqity{\srep{\st''}}$ and $\st'=\sqity{\srep{\st'''}}$ and $\kappa=\kity$ and (1) $\sstepA{\st''}{\st'''}$.
+Rule induction on (b).
+Only rule (k-qity-trans) applies, so (2) $\sofkVX{\st''}{\kity}$.
+Applying the IH to (1) and (2), we have (3) $\sofkVX{\st'''}{\kity}$.
+Apply (k-qity-trans) to (3).
+
+\pfcases{[All rules in Figure \ref{fig:static-dynamics-SL-qity-shared}]} We give an example for forms with 1 and 2 sub-terms.
+All rules with the same number of sub-terms follow the corresponding script, replacing only the forms and rule names.
+
+\begin{quote}
+    \pfcase{(n-q-forall-step)} In this case, we have $\st=\sqity{\iforall{\alpha}{\qity}}$ and $\st'=\sqity{\iforall{\alpha}{\qity'}}$ and (1) $\qtstep{\qity}{\qity'}$.
+Rule induction on (b).
+Only rule (k-qity-forall) applies, so (2) $\sofkVX{\sqity{\qity}}{\kity}$.
+Applying the IH to (1) and (2), we have (3) $\sofkVX{\sqity{\qity'}}{\kity}$.
+Apply (k-qity-forall) to (3).
+
+    \pfcase{(n-q-parr-step-1)} In this case, we have $\st=\sqity{\iparr{\qity_1}{\qity_2}}$ and $\st'=\sqity{\iparr{\qity_1'}{\qity_2}}$ and (1) $\qtstep{\qity_1}{\qity_1'}$.
+Rule induction on (b).
+Only rule (k-qity-parr) applies, so (2) $\sofkVX{\sqity{\qity_1}}{\kity}$ and (3) $\sofkVX{\sqity{\qity_2}}{\kity}$.
+Applying the IH to (1) and (2), we have (4) $\sofkVX{\sqity{\qity_1'}}{\kity}$.
+Apply (k-qity-parr) to (4) and (3).
+
+    \pfcase{(n-q-parr-step-2)} In this case, we have $\st=\sqity{\iparr{\qity_1}{\qity_2}}$ and $\st'=\sqity{\iparr{\qity_1}{\qity_2'}}$ and (1) $\qtstep{\qity_2}{\qity_2'}$.
+Rule induction on (b).
+Only rule (k-qity-parr) applies, so (2) $\sofkVX{\sqity{\qity_1}}{\kity}$ and (3) $\sofkVX{\sqity{\qity_2}}{\kity}$.
+Applying the IH to (1) and (3), we have (4) $\sofkVX{\sqity{\qity_2'}}{\kity}$.
+Apply (k-qity-parr) to (2) and (4).
+\end{quote}
+% \pfcase{(n-q-mu-step)} In this case, we have $\st=\sqity{\imu{t}{\qity}}$ and $\st'=\sqity{\imu{t}{\qity'}}$ and (1) $\qtstep{\qity}{\qity'}$.
+Rule induction on (b).
+Only rule (k-qity-mu) applies, so (2) $\sofkVX{\sqity{\qity}}{\kity}$.
+Applying the IH to (1) and (2), we have (3) $\sofkVX{\sqity{\qity'}}{\kity}$.
+Apply (k-qity-mu) to (3).
+
+% \pfcase{(n-q-prod-step-1)} In this case, we have $\st=\sqity{\iprod{\qity_1}{\qity_2}}$ and $\st'=\sqity{\iprod{\qity_1'}{\qity_2}}$ and (1) $\qtstep{\qity_1}{\qity_1'}$.
+Rule induction on (b).
+Only rule (k-qity-prod) applies, so (2) $\sofkVX{\sqity{\qity_1}}{\kity}$ and (3) $\sofkVX{\sqity{\qity_2}}{\kity}$.
+Applying the IH to (1) and (2), we have (4) $\sofkVX{\sqity{\qity_1'}}{\kity}$.
+Apply (k-qity-prod) to (4) and (3).
+
+% \pfcase{(n-q-prod-step-2)} In this case, we have $\st=\sqity{\iprod{\qity_1}{\qity_2}}$ and $\st'=\sqity{\iprod{\qity_1}{\qity_2'}}$ and (1) $\qtstep{\qity_2}{\qity_2'}$.
+Rule induction on (b).
+Only rule (k-qity-prod) applies, so (2) $\sofkVX{\sqity{\qity_1}}{\kity}$ and (3) $\sofkVX{\sqity{\qity_2}}{\kity}$.
+Applying the IH to (1) and (3), we have (4) $\sofkVX{\sqity{\qity_2'}}{\kity}$.
+Apply (k-qity-prod) to (2) and (4).
+
+% \pfcase{(n-q-sum-step-1)} In this case, we have $\st=\sqity{\isum{\qity_1}{\qity_2}}$ and $\st'=\sqity{\isum{\qity_1'}{\qity_2}}$ and (1) $\qtstep{\qity_1}{\qity_1'}$.
+Rule induction on (b).
+Only rule (k-qity-sum) applies, so (2) $\sofkVX{\sqity{\qity_1}}{\kity}$ and (3) $\sofkVX{\sqity{\qity_2}}{\kity}$.
+Applying the IH to (1) and (2), we have (4) $\sofkVX{\sqity{\qity_1'}}{\kity}$.
+Apply (k-qity-sum) to (4) and (3).
+
+% \pfcase{(n-q-sum-step-2)} In this case, we have $\st=\sqity{\isum{\qity_1}{\qity_2}}$ and $\st'=\sqity{\isum{\qity_1}{\qity_2'}}$ and (1) $\qtstep{\qity_2}{\qity_2'}$.
+Rule induction on (b).
+Only rule (k-qity-sum) applies, so (2) $\sofkVX{\sqity{\qity_1}}{\kity}$ and (3) $\sofkVX{\sqity{\qity_2}}{\kity}$.
+Applying the IH to (1) and (3), we have (4) $\sofkVX{\sqity{\qity_2'}}{\kity}$.
+Apply (k-qity-sum) to (2) and (4).
+
+\pfcase{(n-q-unquote-step)} In this case, we have $\st=\sqitm{\quq{\st''}}$ and $\st'=\sqitm{\quq{\st'''}}$ and $\kappa=\kitm$ and (1) $\sstepA{\st''}{\st'''}$.
+Rule induction on (b).
+Only rule (k-qitm-unquote) applies, so (2) $\sofkVX{\st''}{\kitm}$.
+Applying the IH to (1) and (2), we have (3) $\sofkVX{\st'''}{\kitm}$.
+Apply (k-qitm-unquote) to (3).
+
+\pfcase{(n-q-unquote-elim)} In this case, we have $\st=\sqitm{\quq{\sqitm{\qitm}}}$ and $\st'=\sqitm{\qitm}$ and $\kappa=\kitm$.
+Rule induction on (b).
+Only rule (k-qitm-unquote) applies, so (1) $\sofkVX{\sqitm{\qitm}}{\kitm}$.
+We can conclude by (1).
+
+\pfcase{(n-q-anatrans-step)} In this case, we have $\st=\sqitm{\anatrans{n'}{\st''}}$ and $\st'=\sqitm{\anatrans{n'}{\st'''}}$ and $\kappa=\kitm$ and (1) $\sstepA{\st''}{\st'''}$.
+Rule induction on (b).
+Only rule (k-qitm-anatrans) applies, so (2) $\sofkVX{\st''}{\kitm}$ and (3) $n' < n$.
+Applying the IH to (1) and (2), we have (4) $\sofkVX{\st'''}{\kitm}$.
+Apply (k-qitm-trans) to (3) and (4).
+
+\pfcases{[All rules in Figures \ref{fig:static-dynamics-SL-qitm-shared-1} and \ref{fig:static-dynamics-SL-qitm-shared-2}]} We give an example for forms with 1, 2 and 3 sub-terms.
+All rules with the same number of sub-terms follow the corresponding script, replacing only the forms and rule names.
+
+\begin{quote}
+    \pfcase{(n-q-tabs-step)} In this case, we have $\st=\sqitm{\iLam{\alpha}{\qitm}}$ and $\st'=\sqitm{\iLam{\alpha}{\qitm'}}$ and (1) $\qtstep{\qitm}{\qitm'}$.
+Rule induction on (b).
+Only rule (k-qitm-tabs) applies, so (2) $\sofkVX{\sqitm{\qitm}}{\kitm}$.
+Applying the IH to (1) and (2), we have (3) $\sofkVX{\sqitm{\qitm'}}{\kitm}$.
+Apply (k-qitm-tabs) to (3).
+
+    \pfcase{(n-q-abs-step-1)} In this case, we have $\st=\sqitm{\ilam{\qity_1}{x}{\qitm_2}}$ and $\st'=\sqitm{\ilam{\qity_1'}{x}{\qitm_2}}$ and (1) $\qtstep{\qity_1}{\qity_1'}$.
+Rule induction on (b).
+Only rule (k-qitm-abs) applies, so (2) $\sofkVX{\sqity{\qity_1}}{\kity}$ and (3) $\sofkVX{\sqitm{\qitm_2}}{\kitm}$.
+Applying the IH to (1) and (2), we have (4) $\sofkVX{\sqity{\qity_1'}}{\kity}$.
+Apply (k-qitm-abs) to (4) and (3).
+
+    \pfcase{(n-q-abs-step-2)} In this case, we have $\st=\sqitm{\ilam{\qity_1}{x}{\qitm_2}}$ and $\st'=\sqitm{\ilam{\qity_1}{x}{\qitm_2'}}$ and (1) $\qstep{\qitm_2}{\qitm_2'}$.
+Rule induction on (b).
+Only rule (k-qitm-abs) applies, so (2) $\sofkVX{\sqity{\qity_1}}{\kity}$ and (3) $\sofkVX{\sqitm{\qitm_2}}{\kitm}$.
+Applying the IH to (1) and (3), we have (4) $\sofkVX{\sqitm{\qitm_2'}}{\kitm}$.
+Apply (k-qitm-abs) to (2) and (4).
+
+    \pfcase{(n-q-case-step-1)} In this case, we have $\st=\sqitm{\icase{\qitm}{x}{\qitm_1}{x}{\qitm_2}}$ and $\st'=\sqitm{\icase{\qitm'}{x}{\qitm_1}{x}{\qitm_2}}$ and (1) $\qstep{\qitm}{\qitm'}$.
+Rule induction on (b).
+Only rule (k-qitm-case) applies, so (2) $\sofkVX{\sqitm{\qitm}}{\kitm}$ and (3) $\sofkVX{\sqitm{\qitm_1}}{\kitm}$ and (4) $\sofkVX{\sqitm{\qitm_2}}{\kitm}$.
+Applying the IH to (1) and (2), we have (5) $\sofkVX{\sqitm{\qitm'}}{\kitm}$.
+Apply (k-qitm-case) to (5), (3) and (4).
+
+    \pfcase{(n-q-case-step-2)} In this case, we have $\st=\sqitm{\icase{\qitm}{x}{\qitm_1}{x}{\qitm_2}}$ and $\st'=\sqitm{\icase{\qitm}{x}{\qitm_1'}{x}{\qitm_2}}$ and (1) $\qstep{\qitm_1}{\qitm_1'}$.
+Rule induction on (b).
+Only rule (k-qitm-case) applies, so (2) $\sofkVX{\sqitm{\qitm}}{\kitm}$ and (3) $\sofkVX{\sqitm{\qitm_1}}{\kitm}$ and (4) $\sofkVX{\sqitm{\qitm_2}}{\kitm}$.
+Applying the IH to (1) and (3), we have (5) $\sofkVX{\sqitm{\qitm'}}{\kitm}$.
+Apply (k-qitm-case) to (2), (5) and (4).
+
+    \pfcase{(n-q-case-step-3)} In this case, we have $\st=\sqitm{\icase{\qitm}{x}{\qitm_1}{x}{\qitm_2}}$ and $\st'=\sqitm{\icase{\qitm}{x}{\qitm_1}{x}{\qitm_2'}}$ and (1) $\qstep{\qitm_2}{\qitm_2'}$.
+Rule induction on (b).
+Only rule (k-qitm-case) applies, so (2) $\sofkVX{\sqitm{\qitm}}{\kitm}$ and (3) $\sofkVX{\sqitm{\qitm_1}}{\kitm}$ and (4) $\sofkVX{\sqitm{\qitm_2}}{\kitm}$.
+Applying the IH to (1) and (4), we have (5) $\sofkVX{\sqitm{\qitm_2'}}{\kitm}$.
+Apply (k-qitm-case) to (2), (3) and (5).
+\end{quote}
+% \pfcase{(n-q-ap-step-1)} In this case, we have $\st=\sqitm{\iap{\qitm_1}{\qitm_2}}$ and $\st'=\sqitm{\iap{\qitm_1'}{\qitm_2}}$ and (1) $\qstep{\qitm_1}{\qitm_1'}$.
+Rule induction on (b).
+Only rule (k-qitm-ap) applies, so (2) $\sofkVX{\sqitm{\qitm_1}}{\kitm}$ and (3) $\sofkVX{\sqitm{\qitm_2}}{\kitm}$.
+Applying the IH to (1) and (2), we have (4) $\sofkVX{\sqitm{\qitm_1'}}{\kitm}$.
+Apply (k-qitm-ap) to (4) and (3).
+
+% \pfcase{(n-q-ap-step-2)} In this case, we have $\st=\sqitm{\iap{\qitm_1}{\qitm_2}}$ and $\st'=\sqitm{\iap{\qitm_1}{\qitm_2'}}$ and (1) $\qstep{\qitm_2}{\qitm_2'}$.
+Rule induction on (b).
+Only rule (k-qitm-ap) applies, so (2) $\sofkVX{\sqitm{\qitm_1}}{\kitm}$ and (3) $\sofkVX{\sqitm{\qitm_2}}{\kitm}$.
+Applying the IH to (1) and (3), we have (4) $\sofkVX{\sqitm{\qitm_2'}}{\kitm}$.
+Apply (k-qitm-ap) to (2) and (4).
+
+% \pfcase{(n-q-tap-step-1)} In this case, we have $\st=\sqitm{\iAp{\qity}{\qitm}}$ and $\st'=\sqitm{\iAp{\qity}{\qitm'}}$ and (1) $\qstep{\qitm}{\qitm'}$.
+Rule induction on (b).
+Only rule (k-qitm-tap) applies, so (2) $\sofkVX{\sqity{\qity}}{\kity}$ and (3) $\sofkVX{\sqitm{\qitm}}{\kitm}$.
+Applying the IH to (1) and (3), we have (4) $\sofkVX{\sqitm{\qitm'}}{\kitm}$.
+Apply (k-qitm-tap) to (2) and (4).
+
+% \pfcase{(n-q-tap-step-2)} In this case, we have $\st=\sqitm{\iAp{\qity}{\qitm}}$ and $\st'=\sqitm{\iAp{\qity'}{\qitm}}$ and (1) $\qtstep{\qity}{\qity'}$.
+Rule induction on (b).
+Only rule (k-qitm-tap) applies, so (2) $\sofkVX{\sqity{\qity}}{\kity}$ and (3) $\sofkVX{\sqitm{\qitm}}{\kitm}$.
+Applying the IH to (1) and (2), we have (4) $\sofkVX{\sqity{\qity'}}{\kity}$.
+Apply (k-qitm-tap) to (4) and (3).
+
+% \pfcase{(n-q-fold-step-1)} In this case, we have $\st=\sqitm{\ifold{t}{\qity}{\qitm}}$ and $\st'=\sqitm{\ifold{t}{\qity'}{\qitm}}$ and (1) $\qtstep{\qity}{\qity'}$.
+Rule induction on (b).
+Only rule (k-qitm-fold) applies, so (2) $\sofkVX{\sqity{\qity}}{\kity}$ and (3) $\sofkVX{\sqitm{\qitm}}{\kitm}$.
+Applying the IH to (1) and (2), we have (4) $\sofkVX{\sqity{\qity'}}{\kity}$.
+Apply (k-qitm-fold) to (4) and (3).
+
+% \pfcase{(n-q-fold-step-2)} In this case, we have $\st=\sqitm{\ifold{t}{\qity}{\qitm}}$ and $\st'=\sqitm{\ifold{t}{\qity}{\qitm'}}$ and (1) $\qstep{\qitm}{\qitm'}$.
+Rule induction on (b).
+Only rule (k-qitm-fold) applies, so (2) $\sofkVX{\sqity{\qity}}{\kity}$ and (3) $\sofkVX{\sqitm{\qitm}}{\kitm}$.
+Applying the IH to (1) and (3), we have (4) $\sofkVX{\sqitm{\qitm'}}{\kitm}$.
+Apply (k-qitm-fold) to (2) and (4).
+
+% \pfcase{(n-q-unfold-step)} In this case, we have $\st=\sqitm{\iunfold{\qitm}}$ and $\st'=\sqitm{\iunfold{\qitm'}}$ and (1) $\qtstep{\qitm}{\qitm'}$.
+Rule induction on (b).
+Only rule (k-qitm-unfold) applies, so (2) $\sofkVX{\sqitm{\qitm}}{\kitm}$.
+Applying the IH to (1) and (2), we have (3) $\sofkVX{\sqitm{\qitm'}}{\kitm}$.
+Apply (k-qitm-unfold) to (3).
+
+% \pfcase{(n-q-pair-step-1)} In this case, we have $\st=\sqitm{\ipair{\qitm_1}{\qitm_2}}$ and $\st'=\sqitm{\ipair{\qitm_1'}{\qitm_2}}$ and (1) $\qstep{\qitm_1}{\qitm_1'}$.
+Rule induction on (b).
+Only rule (k-qitm-pair) applies, so (2) $\sofkVX{\sqitm{\qitm_1}}{\kitm}$ and (3) $\sofkVX{\sqitm{\qitm_2}}{\kitm}$.
+Applying the IH to (1) and (2), we have (4) $\sofkVX{\sqitm{\qitm_1'}}{\kitm}$.
+Apply (k-qitm-pair) to (4) and (3).
+
+% \pfcase{(n-q-pair-step-2)} In this case, we have $\st=\sqitm{\ipair{\qitm_1}{\qitm_2}}$ and $\st'=\sqitm{\ipair{\qitm_1}{\qitm_2'}}$ and (1) $\qstep{\qitm_2}{\qitm_2'}$.
+Rule induction on (b).
+Only rule (k-qitm-pair) applies, so (2) $\sofkVX{\sqitm{\qitm_1}}{\kitm}$ and (3) $\sofkVX{\sqitm{\qitm_2}}{\kitm}$.
+Applying the IH to (1) and (3), we have (4) $\sofkVX{\sqitm{\qitm_2'}}{\kitm}$.
+Apply (k-qitm-pair) to (2) and (4).
+
+% \pfcase{(n-q-inl-step-1)} In this case, we have $\st=\sqitm{\iinl{\qity}{\qitm}}$ and $\st'=\sqitm{\iinl{\qity'}{\qitm}}$ and (1) $\qtstep{\qity}{\qity'}$.
+Rule induction on (b).
+Only rule (k-qitm-inl) applies, so (2) $\sofkVX{\sqity{\qity}}{\kity}$ and (3) $\sofkVX{\sqitm{\qitm}}{\kitm}$.
+Applying the IH to (1) and (2), we have (4) $\sofkVX{\sqity{\qity'}}{\kity}$.
+Apply (k-qitm-inl) to (4) and (3).
+
+% \pfcase{(n-q-inl-step-2)} In this case, we have $\st=\sqitm{\iinl{\qity}{\qitm}}$ and $\st'=\sqitm{\iinl{\qity}{\qitm'}}$ and (1) $\qstep{\qitm}{\qitm'}$.
+Rule induction on (b).
+Only rule (k-qitm-inl) applies, so (2) $\sofkVX{\sqity{\qity}}{\kity}$ and (3) $\sofkVX{\sqitm{\qitm}}{\kitm}$.
+Applying the IH to (1) and (3), we have (4) $\sofkVX{\sqitm{\qitm'}}{\kitm}$.
+Apply (k-qitm-inl) to (2) and (4).
+
+% \pfcase{(n-q-inr-step-1)} In this case, we have $\st=\sqitm{\iinr{\qity}{\qitm}}$ and $\st'=\sqitm{\iinr{\qity'}{\qitm}}$ and (1) $\qtstep{\qity}{\qity'}$.
+Rule induction on (b).
+Only rule (k-qitm-inr) applies, so (2) $\sofkVX{\sqity{\qity}}{\kity}$ and (3) $\sofkVX{\sqitm{\qitm}}{\kitm}$.
+Applying the IH to (1) and (2), we have (4) $\sofkVX{\sqity{\qity'}}{\kity}$.
+Apply (k-qitm-inr) to (4) and (3).
+
+% \pfcase{(n-q-inr-step-2)} In this case, we have $\st=\sqitm{\iinr{\qity}{\qitm}}$ and $\st'=\sqitm{\iinr{\qity}{\qitm'}}$ and (1) $\qstep{\qitm}{\qitm'}$.
+Rule induction on (b).
+Only rule (k-qitm-inr) applies, so (2) $\sofkVX{\sqity{\qity}}{\kity}$ and (3) $\sofkVX{\sqitm{\qitm}}{\kitm}$.
+Applying the IH to (1) and (3), we have (4) $\sofkVX{\sqitm{\qitm'}}{\kitm}$.
+Apply (k-qitm-inr) to (2) and (4).
+
+\pfcase{(n-ana-step)} In this case, we have $\st=\sana{n'}{\st''}$ and $\st'=\sana{n'}{\st'''}$ and (1) $\sstepA{\st''}{\st'''}$.
+Rule induction on (b).
+Only rule (k-ana) applies, so (2) $n' < n$ and (3) $\sofkVX{\st''}{\kty}$.
+Applying the IH to (1) and (3), we have (4) $\sofkVX{\st'''}{\kty}$.
+Apply (k-ana) to (2) and (4).
+
+\pfcase{(n-ana-success)} In this case, we have $\st=\sana{n'}{\st''}$ and $\st'=\sqitm{\anatrans{n'}{\st''}}$.
+Rule induction on (b).
+Only rule (k-ana) applies, so (2) $n' < n$ and (3) $\sofkVX{\st''}{\kty}$.
+Apply (k-qitm-anatrans) to (2) and (3).
+
+\pfcase{(n-syn-success)} In this case, we have $\st=\ssyn{n'}$ and (1) $\keyw{nth}[n'](\es)=e$ and (2) $\eanaX{e}{\st''}{\iota}$ and $\st'=(\st'', \sqitm{\syntrans{n'}})$.
+Rule induction on (b).
+Only rule (k-syn) applies, so (3) $n' < n$.
+Applying Theorem \ref{lemma:type-synthesis} to (c), (d), (2), we have that (4) $\sofkVX{\st''}{\kty}$.
+Applying (k-qitm-syntrans) to (3), we have (5) $\sofkVX{\sqitm{\syntrans{n'}}}{\kitm}$.
+Applying (k-prod) to (4) and (5), we have that $\sofkVX{(\st'', \sqitm{\syntrans{n'}})}{\kty\times\kitm}$.
+
+\end{proof}
+
+The final case in the proof of Theorem \ref{thm:static-preservation}, for (n-syn-success), requires that Theorem \ref{lemma:type-synthesis}, shown in Sec.
+\ref{sec:type-synthesis-and-kind-safety} be mutually defined (we give a metric there that strictly decreases, to ensure that the mutual induction is well-founded).
+
+Preservation extends straightforwardly to the multistep judgement $\smanystep{\st}{\argEnv}{\st'}$ in Figure \ref{fig:static-multistep-dynamics}, which is simply the reflexive, transitive closure of the stepping judgement.
+
+\begin{lemma}[Static Multistep Preservation]
+\label{lemma:static-multistep-preservation}
+Letting $\argEnv=\es; \Upsilon; \Phi$, if (a) $\smanystep{\st}{\argEnv}{\st'}$ and (b) $\sofkVX{\st}{\kappa}$ and (c) $\vdash \Phi$ and (d) $\Upsilon~\mathtt{ctx}_\Phi$ then $\sofkVX{\st}{\kappa}$.
+\end{lemma}
+\begin{proof} Rule induction on (a).
+
+\pfcase{(n-multi-refl)} In this case, $\st'=\st$, so the conclusion is assumption (b).
+
+\pfcase{(n-multi-step)} In this case, (1) $\smanystep{\st}{\argEnv}{\st''}$ and (2) $\sstepA{\st''}{\st'}$.
+Applying the IH to (1), (b), (c) and (d), we have that (3) $\sofkVX{\st''}{\kappa}$.
+Applying Theorem \ref{thm:static-preservation} to (b), (3), (c) and (d), we have our conclusion.
+\end{proof}
+
+The normalization judgement is defined as follows:
+
+\begin{definition}[Static Normalization]
+\label{definition:static-normalization}
+$\st \Downarrow_\argEnv \st'$ iff $\smanystep{\st}{\argEnv}{\st'}$ and $\svalA{\st'}$.
+\end{definition}
+
+\begin{corollary}[Static Normalization Preservation]
+\label{lemma:static-normalization-preservation}
+If (a) $\st \Downarrow_\argEnv \st'$ and (b) $\sofkVX{\st}{\kappa}$ and (c) $\vdash \Phi$ and (d) $\Upsilon~\mathtt{ctx}_\Phi$ then $\sofkVX{\st'}{\kappa}$.
+\end{corollary}
+\begin{proof} By Definition \ref{definition:static-normalization} on (a), we have (1) $\smanystep{\st}{\argEnv}{\st'}$.
+By Lemma \ref{lemma:static-multistep-preservation} on (1), (b), (c) and (d), we have our conclusion.
+\end{proof}
+
+\subsection{Stability}
+
+\begin{lemma}[Stability of Kinding]
+\label{lemma:stability-of-kinding}
+If (a) $\tccsigok{\Phi}$ and (b) $\sofkX{\st}{\kappa}$ and (c) $\tccsigok{\Phi'\Phi}$  then $\sofk{\kDelta}{\kGamma}{\Phi'\Phi}{\st}{\kappa}$.
+\end{lemma}
+\begin{proof} By rule induction on (c).
+All cases except the following proceed by  application of the IH to all kinding premises and re-application of the same rule, because they are defined for all $\Phi$ and all premises are checked under the same $\Phi$.
+
+\pfcase{(k-ty-ext)} We have $\st=\sty{\tc}{\sttyidx}$ and (1) $\tcdef{\tc}{\tcsig{\ktyidx}{\chi}}{\theta} \in \Phi$ and (2) $\sofkX{\sttyidx}{\ktyidx}$.
+By the definition of prepending on ordered mappings, we have (3) $\tcdef{\tc}{\tcsig{\ktyidx}{\chi}}{\theta} \in \Phi'\Phi$.
+By the IH on (a), (2) and (c), we have (4) $\sofk{\kDelta}{\kGamma}{\Phi'\Phi}{\sttyidx}{\ktyidx}$.
+Apply (k-ty-ext) on (3) and (4).
+
+\pfcase{(k-tycase-ext)} We have $\st=\stycase{\tc}{\st'}{\sx}{\st_1}{\st_2}$ and (1) $\sofkX{\st'}{\kty}$ and (2) $\tcdef{\tc}{\tcsig{\ktyidx}{\chi}}{\theta} \in \Phi$ and (3) $\sofk{\kDelta}{\kGamma, \sx :: \ktyidx}{\Phi}{\st_1}{\kappa}$ and (4) $\sofkX{\st_2}{\kappa}$.
+
+By the IH on (a), (1) and (c), we have (5) $\sofk{\kDelta}{\kGamma}{\Phi'\Phi}{\st'}{\kty}$.
+
+By the definition of prepending on ordered mappings, we have (6) $\tcdef{\tc}{\tcsig{\ktyidx}{\chi}}{\theta} \in \Phi'\Phi$.
+
+By the IH on (a), (3) and (c), we have (7)  $\sofk{\kDelta}{\kGamma, \sx :: \ktyidx}{\Phi'\Phi}{\st_1}{\kappa}$.
+
+By the IH on (a), (4) and (c), we have (8) $\sofk{\kDelta}{\kGamma}{\Phi'\Phi}{\st_2}{\kappa}$.
+
+Apply (k-tycase-ext) to (5), (6), (7) and (8).
+\end{proof}
+
+\begin{lemma}[Stability of Kinding 2]
+\label{lemma:stability-of-kinding-2}
+If (a) $\tccsigok{\Phi}$ and (b) $\sofkX{\st}{\kappa}$ and (c) $\tccsigok{\Phi\Phi'}$  then $\sofk{\kDelta}{\kGamma}{\Phi\Phi'}{\st}{\kappa}$.
+\end{lemma}
+\begin{proof} By rule induction on (c).
+All cases except the following proceed by  application of the IH to all kinding premises and re-application of the same rule, because they are defined for all $\Phi$ and all premises are checked under the same $\Phi$.
+
+\pfcase{(k-ty-ext)} We have $\st=\sty{\tc}{\sttyidx}$ and (1) $\tcdef{\tc}{\tcsig{\ktyidx}{\chi}}{\theta} \in \Phi$ and (2) $\sofkX{\sttyidx}{\ktyidx}$.
+By the definition of appending ordered mappings, we have (3) $\tcdef{\tc}{\tcsig{\ktyidx}{\chi}}{\theta} \in \Phi\Phi'$.
+By the IH on (a), (2) and (c), we have (4) $\sofk{\kDelta}{\kGamma}{\Phi\Phi'}{\sttyidx}{\ktyidx}$.
+Apply (k-ty-ext) on (3) and (4).
+
+\pfcase{(k-tycase-ext)} We have $\st=\stycase{\tc}{\st'}{\sx}{\st_1}{\st_2}$ and (1) $\sofkX{\st'}{\kty}$ and (2) $\tcdef{\tc}{\tcsig{\ktyidx}{\chi}}{\theta} \in \Phi$ and (3) $\sofk{\kDelta}{\kGamma, \sx :: \ktyidx}{\Phi}{\st_1}{\kappa}$ and (4) $\sofkX{\st_2}{\kappa}$.
+
+By the IH on (a), (1) and (c), we have (5) $\sofk{\kDelta}{\kGamma}{\Phi\Phi'}{\st'}{\kty}$.
+
+By the definition of appending ordered mappings, we have (6) $\tcdef{\tc}{\tcsig{\ktyidx}{\chi}}{\theta} \in \Phi\Phi'$.
+
+By the IH on (a), (3) and (c), we have (7)  $\sofk{\kDelta}{\kGamma, \sx :: \ktyidx}{\Phi\Phi'}{\st_1}{\kappa}$.
+
+By the IH on (a), (4) and (c), we have (8) $\sofk{\kDelta}{\kGamma}{\Phi\Phi'}{\st_2}{\kappa}$.
+
+Apply (k-tycase-ext) to (5), (6), (7) and (8). \end{proof}
+
+\begin{corollary}[Stability of Kinding 3]
+\label{lemma:stability-of-kinding-3}
+If (a) $\vdash \Phi$ and (b) $\sofkX{\st}{\kappa}$ and (c) $\vdash \Phi\Phi'$ then $\sofk{\kDelta}{\kGamma}{\Phi\Phi'}{\st}{\kappa}$.
+\end{corollary}
+\begin{proof}
+Applying Lemma \ref{lemma:well-defined-contexts-have-well-formed-signatures} on (a), we have (1) $\tccsigok{\Phi}$.
+
+Applying Lemma \ref{lemma:well-defined-contexts-have-well-formed-signatures} on (c), we have (2) $\tccsigok{\Phi\Phi'}$.
+
+Apply Lemma \ref{lemma:stability-of-kinding-2} to (1), (b) and (2).
+\end{proof}
+
+We can now give corresponding stability lemmas about tycon context extension.
+
+\begin{lemma}[Stability of Opcon Structure Validity]
+\label{lemma:stability-of-opcon-structure-validity}
+If (a) $\tccsigok{\Phi}$ and (b) $\vdash_\Phi \omega \sim \psi$ and (c)  $\tccsigok{\Phi'\Phi}$ then $\vdash_{\Phi'\Phi} \omega \sim \psi$. 
+\end{lemma}
+\begin{proof} Rule induction on (b).
+
+\pfcase{(ocstruct-intro)} We have $\omega=\tcstructn{\stx{def}}$ and (1) $\emptyset \vdash \klitidx$ and (2) $
+    \sofkz{\emptyset}{\emptyset}{\Phi}{\stx{def}}{\ktyidx \rightarrow \klitidx \rightarrow \kargs \rightarrow \kitm}$.
+
+    By Lemma \ref{lemma:stability-of-kinding} on (a), (2) and (c), we have (3) $
+    \sofkz{\emptyset}{\emptyset}{\Phi'\Phi}{\stx{def}}{\ktyidx \rightarrow \klitidx \rightarrow \kargs \rightarrow \kitm}$.
+
+    Apply (ocstruct-intro) to (1) and (3).
+
+\pfcase{(ocstruct-targ)} We have $\omega=\tcstructc{\omega'}{op}{\stx{def}}$ and $\psi=\tcsig{\ktyidx}{\chi; \opsig{\opname{op}}{\klitidx}}$ and (1) $\vdash_\Phi \omega' \sim \tcsig{\ktyidx}{\chi}$ and (2) $\emptyset \vdash \klitidx$ and (3) $\sofkn{\emptyset}{\emptyset}{\Phi}{0}{\stx{def}}{\ktyidx \rightarrow \klitidx \rightarrow \kargs \rightarrow (\kty \times \kitm)}$.
+
+By the IH on (a), (1) and (c), we have (4) $\vdash_{\Phi'\Phi} \omega' \sim \tcsig{\ktyidx}{\chi}$.
+
+By Lemma \ref{lemma:stability-of-kinding} on (a), (3) and (c), we have (5) $\sofkn{\emptyset}{\emptyset}{\Phi'\Phi}{0}{\stx{def}}{\ktyidx \rightarrow \klitidx \rightarrow \kargs \rightarrow (\kty \times \kitm)}$.
+
+Apply (ocstruct-targ) to (4), (2) and (5).
+\end{proof}
+
+\begin{lemma}[Stability of Opcon Structure Validity 2]
+\label{lemma:stability-of-opcon-structure-validity-2}
+If (a) $\tccsigok{\Phi}$ and (b) $\vdash_\Phi \omega \sim \psi$ and (c)  $\tccsigok{\Phi\Phi'}$ then $\vdash_{\Phi\Phi'} \omega \sim \psi$. 
+\end{lemma}
+\begin{proof} Rule induction on (b).
+
+\pfcase{(ocstruct-intro)} We have $\omega=\tcstructn{\stx{def}}$ and (1) $\emptyset \vdash \klitidx$ and (2) $
+    \sofkz{\emptyset}{\emptyset}{\Phi}{\stx{def}}{\ktyidx \rightarrow \klitidx \rightarrow \kargs \rightarrow \kitm}$.
+
+    By Lemma \ref{lemma:stability-of-kinding-2} on (a), (2) and (c), we have (3) $
+    \sofkz{\emptyset}{\emptyset}{\Phi\Phi'}{\stx{def}}{\ktyidx \rightarrow \klitidx \rightarrow \kargs \rightarrow \kitm}$.
+
+    Apply (ocstruct-intro) to (1) and (3).
+
+\pfcase{(ocstruct-targ)} We have $\omega=\tcstructc{\omega'}{op}{\stx{def}}$ and $\psi=\tcsig{\ktyidx}{\chi; \opsig{\opname{op}}{\klitidx}}$ and (1) $\vdash_\Phi \omega' \sim \tcsig{\ktyidx}{\chi}$ and (2) $\emptyset \vdash \klitidx$ and (3) $\sofkn{\emptyset}{\emptyset}{\Phi}{0}{\stx{def}}{\ktyidx \rightarrow \klitidx \rightarrow \kargs \rightarrow (\kty \times \kitm)}$.
+
+By the IH on (a), (1) and (c), we have (4) $\vdash_{\Phi\Phi'} \omega' \sim \tcsig{\ktyidx}{\chi}$.
+
+By Lemma \ref{lemma:stability-of-kinding} on (a), (3) and (c), we have (5) $\sofkn{\emptyset}{\emptyset}{\Phi\Phi'}{0}{\stx{def}}{\ktyidx \rightarrow \klitidx \rightarrow \kargs \rightarrow (\kty \times \kitm)}$.
+
+Apply (ocstruct-targ) to (4), (2) and (5).
+\end{proof}
+
+\begin{corollary}[Stability of Opcon Structure Validity 3]
+\label{lemma:stability-of-opcon-structure-validity-3}
+If (a) $\vdash \Phi$ and (b) $\vdash_\Phi \omega \sim \psi$ and (c) $\vdash \Phi\Phi'$ then $\vdash_{\Phi\Phi'} \omega \sim \psi$.
+\end{corollary}
+\begin{proof} Applying Lemma \ref{lemma:well-defined-contexts-have-well-formed-signatures} to (a), we have (1) $\tccsigok{\Phi}$.
+
+Applying Lemma \ref{lemma:well-defined-contexts-have-well-formed-signatures} to (c), we have (2) $\tccsigok{\Phi\Phi'}$.
+
+Apply Lemma \ref{lemma:stability-of-opcon-structure-validity-2} to (1), (b) and (2).
+\end{proof}
+
+\begin{lemma}[Tycon Context Extension]
+If (a) $\vdash \Phi$ and (b) $\vdash \Phi'$ and (c) $\text{dom}(\Phi) \cap \text{dom}(\Phi') = \emptyset$ then $\vdash \Phi\Phi'$.
+\end{lemma}
+\begin{proof} Rule induction on (b).
+
+\pfcase{(tcc-emp)} Here, $\Phi'=\emptyset$, so $\Phi\Phi'=\Phi$ and $\vdash \Phi$ by (a).
+
+\pfcase{(tcc-ext)} Here, $\Phi'=\Phi'', \tcdef{\tc}{\tcsig{\ktyidx}{\chi}}{\tcstruct{\stx{schema}}{\omega}}$ and (1) $\vdash \Phi''$ and (2) $\keq{}{\ktyidx}$ and (3) $
+    \sofkz{\emptyset}{\emptyset}{\Phi''}{\stx{schema}}{\ktyidx \rightarrow \kity}$ and (4) $
+        \vdash_{\Phi'',  \tcdef{\tc}{\tcsig{\ktyidx}{\chi}}{\tcstruct{\stx{schema}}{\omega}}} \omega \sim \tcsig{\ktyidx}{\chi}$.
+
+        By the fact that $\Phi''$ is a prefix of $\Phi'$ and (c), we have that (5) $\text{dom}(\Phi) \cap \text{dom}(\Phi'')$.
+
+        By the IH on (a), (1) and (5), we have (6) $\vdash \Phi\Phi''$.
+
+        By Lemma \ref{lemma:well-defined-contexts-have-well-formed-signatures} on (6), we have (7) $\tccsigok{\Phi\Phi''}$.
+
+        By Lemma \ref{lemma:well-defined-contexts-have-well-formed-signatures} on (1), we have (8) $\tccsigok{\Phi''}$.
+
+        By Lemma \ref{lemma:stability-of-kinding} on (8), (3) and (7), we have (9) $\sofkz{\emptyset}{\emptyset}{\Phi\Phi''}{\stx{schema}}{\ktyidx \rightarrow \kity}$.
+
+        By Lemma \ref{lemma:chi-ok} on (4), we have (10) $\vdash \chi$.
+
+        By (tcsig-ok) on (2) and (10), we have (11) $\vdash \tcsig{\ktyidx}{\chi}$.
+
+        By (tcc-sigok-ext) on (7) and (11), we have (12) $\tccsigok{\Phi\Phi'', \tcdef{\tc}{\tcsig{\ktyidx}{\chi}}{\tcstruct{\stx{schema}}{\omega}}}$.
+
+        By (tcc-sigok-ext) on (8) and (11), we have (13) $\tccsigok{\Phi'', \tcdef{\tc}{\tcsig{\ktyidx}{\chi}}{\tcstruct{\stx{schema}}{\omega}}}$.
+
+        By Lemma \ref{lemma:stability-of-opcon-structure-validity} on (13), (4) and (12), we have (14) $
+        \vdash_{\Phi\Phi'',  \tcdef{\tc}{\tcsig{\ktyidx}{\chi}}{\tcstruct{\stx{schema}}{\omega}}} \omega \sim \tcsig{\ktyidx}{\chi}$.
+
+        By (tcc-ext) on (6), (2), (9) and (14), we have our conclusion. 
+\end{proof}
+
+\newpage
 \section{External Language}
-\subsection{Additional Desugarings}
+\subsection{Syntax}
+We now turn to the external language.
+The abstract syntax of the external language is given in Figure \ref{syntax-EL} and Figures \ref{fig:example-intro-desugarings} and \ref{fig:example-targeted-desugarings} give various desugarings.
+Although we do not detail it here, we expect a technique based on \cite{TSLs} could be introduced to allow new desugarings to be introduced composably.
+
+Argument lists, $\es$, are assumed to be defined as metatheoretic lists, so we assume standard operations like computing the length, written $|\es|=n$, or retrieving an element, written $\keyw{nth}[n](\es)=e$, are already defined and obey standard metatheoretic properties.
+
 \subsection{Typing}
-\subsection{Proof of Regular String Soundness Tycon Invariant}
-
-\newpage
+The typing rules are given in Figure \ref{fig:typing}. We will cover their premises in the next several subsections, then return to the metatheory after that.
+
+\subsection{Argument Interfaces}
+
+\begin{definition}[Argument Interface Kind]
+The kind abbreviated $\karg$ is defined as\[\karg := (\kty \rightarrow \kitm) \times (\kunit \rightarrow \kty\times\kitm)\]
+\end{definition}
+
+\begin{lemma}[Argument Interface Kind is a Kind] 
+\label{lemma:argument-interface-kind-is-kind}
+We have that $\emptyset \vdash \karg$.
+\end{lemma}
+\begin{proof} We apply the following sequences of rules to construct the derivation in a bottom up, left-to-right manner:
+
+Apply (kf-prod). Apply (kf-arrow). Apply (kf-ty). Apply (kf-itm). Apply (kf-arrow). Apply (kf-unit). Apply (kf-prod). Apply (kf-ty). Apply (kf-itm).
+\end{proof}
+
+The argument interfaces that populate the list provided to opcon definitions is derived from the argument list $\es$ by the judgement $\keyw{args}(n)= \stx{args}$, where the $n$ we use is the length of the argument list (including the target argument when applicable).
+This judgement is defined in Figure \ref{fig:arg-interface-list} with an auxiliary judgement $\keyw{args}(n, m)=\stx{args}$, where $m$ is used to count up to $n$ so that the argument indices are in the correct order.
+
+\begin{lemma}[Argument Interface Lists (Auxiliary)]
+\label{lemma:argument-interface-lists-aux}
+If $\keyw{args}(n, m)=\stx{args}$ then (i) $\svalA{\stx{args}}$ and (ii) $\sofkn{\emptyset}{\emptyset}{\Phi}{n+m}{\stx{args}}{\kargs}$.
+\end{lemma}
+\begin{proof} Rule induction on the assumption.
+
+\pfcase{(mk-arg-interface-nil)} In this case, $n=0$ and thus $n+m=m$, and $\stx{args}=\keyw{nil}[\karg]$.
+Applying (n-nil-val), we have (i) $\svalA{\keyw{nil}[\karg]}$.
+By Lemma \ref{lemma:argument-interface-kind-is-kind}, we have (1) $\emptyset \vdash \karg$.
+Applying (k-nil) to (1), we have (ii) $\sofkn{\emptyset}{\emptyset}{\Phi}{m}{\keyw{nil}[\karg]}{\klist{\karg}}$.
+
+\pfcase{(mk-arg-interface-cons)} In this case, $n=n'+1$ and (1) $\keyw{args}(n', m+1)=\stx{args}'$ and $$\stx{args}=\keyw{cons}((\slam{\svar{ty}}{\kty}{\sana{m}{\svar{ty}}}, \slam{\kunit}{\_}{\ssyn{m}}); \stx{args}')$$
+
+Applying the IH to (1), we have that (2) $\svalA{\stx{args}'}$ and (3) $\sofkn{\emptyset}{\emptyset}{\Phi}{n'+(m+1)}{\stx{args}'}{\kargs}$.
+Noting that $(n'+1)+m=n'+(m+1)$ by the usual properties of arithmetic, we thus have that (4) $\sofkn{\emptyset}{\emptyset}{\Phi}{(n'+1)+m}{\stx{args}'}{\kargs}$.
+
+Applying (kf-ty), we have (5) $\emptyset \vdash \kty$.
+
+Applying (k-var), we have that (6) $\sofkn{\emptyset}{\emptyset, \svar{ty} :: \kty}{\Phi}{(n' + 1) + m}{\svar{ty}}{\kty}$.
+
+We have that (7) $m < (n' + 1) + m$ by the usual properties of arithmetic inequalities.
+
+Applying (k-ana) to (7) and (6), we have (8) $\sofkn{\emptyset}{\emptyset, \svar{ty} :: \kty}{\Phi}{(n' + 1) + m}{\sana{m}{\svar{ty}}}{\kitm}$.
+
+Applying (k-abs) to (5) and (8), we have (9) $\sofkn{\emptyset}{\emptyset}{\Phi}{(n'+1)+m}{\slam{\kty}{\svar{ty}}{\sana{m}{\svar{ty}}}}{\karrow{\kty}{\kitm}}$.
+
+Applying (k-syn) to (7), we have (10) $\sofkn{\emptyset}{\emptyset, \_ :: \kunit}{\Phi}{(n' + 1) + m}{\ssyn{m}}{\kty\times\kitm}$.
+
+Applying (kf-unit), we have (11) $\emptyset \vdash \kunit$.
+
+Applying (k-abs) to (11) and (1), we have (12) $\sofkn{\emptyset}{\emptyset}{\Phi}{(n' + 1) + m}{\slam{\kunit}{\_}{\ssyn{m}}}{\karrow{\kunit}{\kty\times\kitm}}$.
+
+Applying (k-pair) to (9) and (12), we have (13) $$\sofkn{\emptyset}{\emptyset}{\Phi}{(n' + 1) + m}{(\slam{\svar{ty}}{\kty}{\sana{m}{\svar{ty}}}, \slam{\kunit}{\_}{\ssyn{m}})}{\karg}$$
+
+Applying (k-cons) to (12) and (4), we have (ii)  $$\sofkn{\emptyset}{\emptyset}{\Phi}{(n' + 1) + m}{\keyw{cons}((\slam{\svar{ty}}{\kty}{\sana{m}{\svar{ty}}}, \slam{\kunit}{\_}{\ssyn{m}}); \stx{args}')}{\kargs}$$
+
+Applying (n-abs-val), we have (14) $\svalA{\slam{\svar{ty}}{\kty}{\sana{m}{\svar{ty}}}}$.
+
+Applying (n-abs-val), we have (15) $\svalA{\slam{\kunit}{\_}{\ssyn{m}}}$.
+
+Applying (n-pair-val) to (14) and (15), we have (16) $\svalA{(\slam{\svar{ty}}{\kty}{\sana{m}{\svar{ty}}}, \slam{\kunit}{\_}{\ssyn{m}})}$.
+
+Applying (n-cons-val) to (16) and (2), we have (i) $$\svalA{\keyw{cons}((\slam{\svar{ty}}{\kty}{\sana{m}{\svar{ty}}}, \slam{\kunit}{\_}{\ssyn{m}}); \stx{args}')}$$
+\end{proof}
+
+\begin{lemma}[Argument Interface Lists]
+\label{lemma:argument-interface-lists}
+If $\keyw{args}(n)=\stx{args}$ then (i) $\svalA{\stx{args}}$ and (ii) $\sofkn{\emptyset}{\emptyset}{\Phi}{n}{\stx{args}}{\kargs}$.
+\end{lemma}
+\begin{proof} Rule induction on the assumption.
+There is only one rule, (mk-arg-interface), so we have that (1) $\keyw{args}(n, 0)=\stx{args}$.
+Applying Lemma \ref{lemma:argument-interface-lists-aux} to (1) and noting that $n+0=n$ by definition, we have (i) $\svalA{\stx{args}}$ and (ii) $\sofkn{\emptyset}{\emptyset}{\Phi}{n}{\stx{args}}{\kargs}$.
+\end{proof}
+
+\subsection{Type Synthesis and Static Preservation}\label{sec:type-synthesis-and-kind-safety}
+
+To prove the type synthesis theorem below, we need a notion of external typing context validity, defined by the judgement $\Upsilon~\mathtt{ctx}_{\Phi}$ in Figure \ref{fig:typing-context-well-formedness}.
+It simply ensures that all bindings in $\Upsilon$ map to types.
+
+\begin{lemma}[Typing Context Validity]
+\label{lemma:typing-context-validity}
+If (a) $\Upsilon~\mathtt{ctx}_{\Phi}$ and (b) $x : \st \in \Upsilon$ then $\istype{\st}{\Phi}$.
+\end{lemma}
+\begin{proof} Rule induction on (a) and (b).
+
+\pfcase{(tctx-ok-emp)} Does not apply by (b).
+
+\pfcase{(tctx-ok-ext)} In this case, $\Upsilon=\Upsilon', x' : \st'$ and (1) $\Upsilon~\mathtt{ctx}_{\Phi}'$ and (2) $\istype{\st'}{\Phi}$. We have two cases by (b):
+
+\begin{quote}
+    \pfcase{$x = x'$ and $\st=\st'$} In this case, we have our conclusion by (2).
+
+    \pfcase{$x \neq x'$ and (3) $x : \st \in \Upsilon'$} In this case, we apply our IH to (1) and (3) to conclude.
+\end{quote}
+\end{proof}
+
+\begin{theorem}[Type Synthesis]
+\label{lemma:type-synthesis}
+If (a) $\vdash \Phi$ and (b) $\Upsilon~\mathtt{ctx}_{\Phi}$ and (c) $\esynX{e}{\st}{\iota}$ then $\istype{\st}{\Phi}$.
+\end{theorem}
+\begin{proof}Rule induction on (c).
+
+\pfcase{(ascribe)} In this case, we have $e=e' : \st'$ and (1) $\sofkVXz{\st'}{\kty}$ and (2) $\st' \Downarrow_{\cdot;\emptyset;\Phi} \st$.
+By Corollary \ref{lemma:static-normalization-preservation} on (2), (1), (a) and (b), we have (3) $\sofkVXz{\st}{\kty}$.
+By Definition \ref{definition:static-normalization} on (2), we have (4) $\sval{\st}{\cdot;\emptyset;\Phi}$.
+Apply Definition \ref{def:types} to (3) and (4).
+
+\pfcase{(var)} In this case, we have (1) $x : \st \in \Upsilon$. Apply Lemma \ref{lemma:typing-context-validity} to (a) and (1).
+
+\pfcase{(syn-lam)} In this case, we have $e=\elam{\st_1}{x}{e'}$ and $\st=\sty{\rightharpoonup}{(\st_1', \st_2)}$ and (1) $\sofkVXz{\st_1}{\kty}$ and (2) $\st_1 \Downarrow_{\cdot;\emptyset;\Phi} \st_1'$ and (3) $\esyn{\Upsilon, x : \st_1'}{\Phi}{e'}{\st_2}{\iota'}$.
+
+By Corollary \ref{lemma:static-normalization-preservation} on (2), (1), (a) and (b), we have (3) $\sofkVXz{\st_1'}{\kty}$.
+By Definition \ref{definition:static-normalization} on (2), we have (4) $\sval{\st_1'}{\cdot;\emptyset;\Phi}$.
+By Definition \ref{def:types} on (3) and (4), we have (5) $\istype{\st_1'}{\Phi}$.
+
+By (tctx-ok-ext) to (b) and (5), we have (6) $\Upsilon, x : \st_1'~\mathtt{ctx}_\Phi$.
+
+Applying the IH to (a), (6) and (3), we have (7) $\istype{\st_2}{\Phi}$ and so by Definition \ref{def:types} on (7), we have (8) $\sofkVXz{\st_2}{\kty}$ and (9) $\sval{\st_2}{\cdot;\emptyset;\Phi}$.
+
+Applying (k-pair) to (3) and (8), we have (10) $\sofkVXz{(\st_1', \st_2)}{\kty\times\kty}$.
+
+Applying (k-ty-parr) to (10), we have (11) $\sofkVXz{\sty{\rightharpoonup}{(\st_1', \st_2)}}{\kty}$.
+
+Applying (n-pair-val) to (4) and (9), we have (12) $\sval{(\st_1', \st_2)}{\cdot;\emptyset;\Phi}$.
+
+Applying (n-ty-val) to (12), we have $\sval{\sty{\rightharpoonup}{(\st_1', \st_2)}}{\cdot;\emptyset;\Phi}$.
+
+Apply Definition \ref{def:types} to (11) and (13).
+
+\pfcase{(syn-ap)} In this case, $e=\eap{e_1}{e_2}$ and (1) $\esynX{e_1}{\sty{\rightharpoonup}{(\st_1, \st)}}{\iota_1}$.
+Applying the IH to (a), (b) and (1), we have (2) $\istype{\sty{\rightharpoonup}{(\st_1, \st)}}{\Phi}$.
+By Definition \ref{def:types} on (2), we have (3) $\sofkVXz{\sty{\rightharpoonup}{(\st_1, \st)}}{\kty}$ and (4) $\svalA{\sty{\rightharpoonup}{(\st_1, \st)}}$.
+By Lemma \ref{lemma:static-canonical-forms} on (4) and (3), we have $\istype{\st}{\Phi}$.
+
+\pfcase{(syn-targ)} In this case, $e=\etarg{\opname{op}}{\stx{tmidx}}{e_\text{targ}}{\es}$ and \begin{enumerate}[(1)]
+\item $\esynX{e_\text{targ}}{\sty{\tc}{\sttyidx}}{\iota_\text{targ}}$ and 
+\item $\tcdef{\tc}{\tcsig{\ktyidx}{\chi}}{\tcstruct{\stx{schema}}{\omega}} \in \Phi$ and
+\item $\opsig{\opname{op}}{\klitidx} \in \chi$ and
+\item $\sofkz{\emptyset}{\emptyset}{\Phi}{\stmidx}{\klitidx}$ and
+\item $\keyw{syn~}\opname{op}={\stx{def}} \in \omega$ and
+\item $|e_\text{targ}; \es| = n$ and
+\item $\keyw{args}(n)=\stx{args}$ and
+\item $\stx{def}(\sttyidx)(\stmidx)(\stx{args}) \Downarrow_{(e_\text{targ}; \es);\Upsilon;\Phi} (\st, \sqitm{\qitm})$.
+\end{enumerate}
+
+Applying the IH to (a), (b) and (1), we have (9) $\istype{\sty{\tc}{\sttyidx}}{\Phi}$.
+By Lemma \ref{lemma:static-canonical-forms}.7 on (9), we have (10) $\sofkVXz{\sttyidx}{\ktyidx}$.
+Applying Lemma \ref{lemma:weakening-of-argument-bounds} to (10), we have (11) $\sofkVX{\sttyidx}{\ktyidx}$.
+
+Applying Lemma \ref{lemma:tycon-context-validity} to (2), we have that there is a prefix $\Phi'$ of $\Phi$ such that (12) $\vdash \Phi'$ and (13) $\vdash \Phi', \tcdef{\tc}{\tcsig{\ktyidx}{\chi}}{\tcstruct{\stx{schema}}{\omega}}$ and (14) $\vdash_{\Phi', \tcdef{\tc}{\tcsig{\ktyidx}{\chi}}{\tcstruct{\stx{schema}}{\omega}}} \omega \sim \tcsig{\ktyidx}{\chi}$.
+
+Applying Lemma \ref{lemma:well-defined-contexts-have-well-formed-signatures} to (12), we have (15) $\tccsigok{\Phi', \tcdef{\tc}{\tcsig{\ktyidx}{\chi}}{\tcstruct{\stx{schema}}{\omega}}}$.
+Applying Lemma \ref{lemma:well-defined-contexts-have-well-formed-signatures} to (a), taking by the definition of prefixing that $\Phi=(\Phi', \tcdef{\tc}{\tcsig{\ktyidx}{\chi}}{\tcstruct{\stx{schema}}{\omega}})\Phi''$, we have (16) $\tccsigok{(\Phi', \tcdef{\tc}{\tcsig{\ktyidx}{\chi}}{\tcstruct{\stx{schema}}{\omega}})\Phi''}$.
+Applying Lemma \ref{lemma:stability-of-opcon-structure-validity} to (15), (14) and (16), we have that (17) $\vdash_{\Phi} \omega \sim \tcsig{\ktyidx}{\chi}$.
+
+Applying Lemma \ref{lemma:targeted-opcon-validity} to (17) and (5), we have that (18) $\sofkz{\emptyset}{\emptyset}{\Phi}{\stx{def}}{\ktyidx \rightarrow \klitidx \rightarrow \kargs \rightarrow (\kty \times \kitm)}$.
+Applying Lemma \ref{lemma:weakening-of-argument-bounds} to (18), we have (19) $\sofk{\emptyset}{\emptyset}{\Phi}{\stx{def}}{\ktyidx \rightarrow \klitidx \rightarrow \kargs \rightarrow (\kty \times \kitm)}$.
+
+Applying Lemma \ref{lemma:argument-interface-lists} to (7), we have that (20) $\sofkVX{\stx{args}}{\kargs}$.
+
+Applying (k-ap) three times with (19), (11), (4) and (20), we have that (21) $\sofkVX{\stx{def}(\sttyidx)(\stx{tmidx})(\stx{args})}{\kty\times\kitm}$.
+
+Applying Corollary \ref{lemma:static-normalization-preservation} (see note on well-foundedness below) to (8), (21), (a) and (b), we have that (22) $\sofkVX{(\st, \sqitm{\qitm})}{\kty\times\kitm}$.
+By Definition \ref{definition:static-normalization} on (8), we have (23) $\sval{(\st, \sqitm{\qitm})}{(e_\text{targ}; \es); \Upsilon; \Phi}$.
+
+Applying Lemma \ref{lemma:static-canonical-forms}.5 to (23) and (22), we have that (24) $\sofkVX{\st}{\kty}$ and (25) $\sval{\st}{(e_\text{targ}; \es); \Upsilon; \Phi}$.
+Apply Definition \ref{def:types} to (24) and (25).
+
+\pfcase{(syn-targ-other)} In this case, we have that $\st=\etarg{\opname{op}}{(\st, \sqitm{\qitm})}{e_\text{targ}}{\es}$ and (1) $\istype{\st}{\Phi}$, so we can conclude by (1).
+\end{proof}
+The mutual induction is well-founded because the total number of intro and targeted terms that remain to be typechecked strictly decreases. In particular, we define the following metrics:
+
+\newcommand{\esize}[1]{\|#1\|}
+
+\begin{definition}[Extension Operation Count] The extension operation count, written $\esize{e}=n$ on external terms and $\esize{\es}=n$ on argument lists, is mutually defined as follows:
+\[\begin{array}{rcl}
+\esize{x} & = & 0\\
+\esize{\eanalam{x}{e}} & = & \esize{e}\\
+\esize{\elam{\st}{x}{e}} & = & \esize{e}\\
+\esize{\eap{e_1}{e_2}} & = & \esize{e_1} + \esize{e_2}\\
+\esize{\efix{x}{e}} & = & \esize{e}\\
+\esize{e : \st} & = & \esize{e}\\
+\esize{\eintro{\st}{\es}} & = & \esize{\es} + 1\\
+\esize{\etarg{\opname{op}}{\st}{e_\text{targ}}{\es}} & = & \esize{e_\text{targ}} + \esize{\es} + 1\\
+\\
+\esize{\cdot} & = & 0\\
+\esize{\es; e} & = & \esize{\es} + \esize{e}
+\end{array}\]
+\end{definition}
+
+In the proof of Theorem \ref{lemma:type-synthesis} for an external term $e$ such that $\esize{e}=n$, Corollary \ref{lemma:static-normalization-preservation} (of Theorem \ref{thm:static-preservation}) is invoked only when $e=\etarg{\opname{op}}{\st'}{e_\text{targ}}{\es'}$ with an argument list $\es=e_\text{targ}; \es'$.
+By definition, $\esize{e}=\esize{e_\text{targ}} + \esize{\es'} + 1$ and $\esize{\es}=\esize{e_\text{targ}} + \esize{\es'}$, so $\esize{\es} < \esize{e}$.
+
+In the proof of Theorem \ref{thm:static-preservation} with an argument list $\es$ such that $\esize{\es}=n$, Theorem \ref{lemma:type-synthesis} is invoked only on a term $e$ extracted from $\es$, so by definition $\esize{e} \leq \esize{\es}$.
+
+This metric strictly decreases every time we pass into Theorem \ref{thm:static-preservation} and does not increase in passing back to Theorem \ref{lemma:type-synthesis}, so mutual induction is well-founded.
+
+\subsection{Type and Typing Context Translations}
+The type translation judgement $\tytransX{\st}{\tau}$ specified in Figure \ref{fig:type-translation} simply checks that $\st$ is a type, then generates any selectively abstracted type translation and applies the type substitution it implies, discussed below.
+
+\begin{lemma}[Types with Translations are Types]
+\label{lemma:types-with-translations-are-types}
+If $\tytransX{\st}{\tau}$ then $\istype{\st}{\Phi}$.
+\end{lemma}
+\begin{proof} Rule induction on the assumption. The conclusion is the first premise of the only rule.\end{proof}
+
+Because typing contexts map variables to types, we need an analagous context translation judgement, $\ctxtransX{\Upsilon}{\Gamma}$, specified in Figure \ref{fig:typing-context--translation}.
+
+\begin{lemma}[Typing Contexts with Translations are Valid Typing Contexts]
+If $\ctxtransX{\Upsilon}{\Gamma}$ then $\vdash_\Phi \Upsilon$.
+\end{lemma}
+\begin{proof} Rule induction on the assumption.
+
+\pfcase{(ctx-trans-emp)} Apply (tctx-ok-emp).
+
+\pfcase{(ctx-trans-ext)} We have that $\Upsilon=\Upsilon', x : \st$ and (1) $\ctxtransX{\Upsilon'}{\Gamma'}$ and (2) $\tytransX{\st}{\tau}$.
+
+Applying the IH to (1), we have that (3) $\vdash_\Phi \Upsilon'$.
+Applying Lemma \ref{lemma:types-with-translations-are-types} to (2), we have (4) $\istype{\st}{\Phi}$.
+Apply (tctx-ok-ext) to (3) and (4).
+\end{proof}
+
+\subsubsection{Selectively Abstracted Type Translations}
+The rules for generating selectively abstracted type translations are given in Figure \ref{fig:selectively-abstracted-type-translations}.
+They make use of a type translation store, $\memD$.
+For our metatheory, we need a notion of a valid type translation store under tycon context $\Phi$ for delegated tycon $c$, written $\tytsvX{\memD}$ and specified in Figure \ref{fig:type-translation-store-validity}.
+Each type translation store implies a type variable substitution and internal type formation context, $\Delta$, by the internalization judgement $\memD \leadsto \delta : \Delta$, specified in Figure \ref{fig:type-translation-store-Internalization}.
+
+\begin{lemma}[Type Translation Store Validity]
+\label{lemma:type-translation-store-validitiy}
+If (a) $\tytsvX{\memD}$ and (b) $\sty{c'}{\sttyidx} \leftrightarrow \tau/\alpha \in \memD$ then (i) $c\neq c'$ and (ii) $\tytransX{\sty{c'}{\sttyidx}}{\tau}$ and (iii) $\emptyset \vdash \tau$.
+\end{lemma}
+\begin{proof} Rule induction on (a) and (b).
+
+\pfcase{(tyts-ok-emp)} Does not apply by (b).
+
+\pfcase{(tyts-ok-ext)} We have two cases by (b):
+\begin{quote}
+    \pfcase{$\memD=\memD', \sty{c'}{\sttyidx}$ and (1) $c\neq c'$ and (2) $\tytransX{\sty{c'}{\sttyidx}}{\tau}$} and (3) $\emptyset \vdash \tau$. We can conclude with (1) and (2).
+
+    \pfcase{$\memD=\memD', \sty{c''}{\sttyidx'} \leftrightarrow \tau/\alpha$ for $c'\neq c''$ and (1) $\sty{c'}{\sttyidx} \leftrightarrow \tau/\alpha \in \memD$ and (2) $\tytsvX{\memD'}$} Apply the IH to  (2) and (1).
+\end{quote}
+\end{proof}
+
+\begin{lemma}[Selective Type Translation Abstraction Caching]
+\label{lemma:type-translation-stores-are-idempotent}
+If (a) $\sofkVX{\sqity{\qity}}{\kity}$ and (b) $\sval{\sqity{\qity}}{\es;\Upsilon;\Phi}$ and (c) $\vdash \Phi$ and (d) $\tytsvX{\memD_1}$ and (e) $\tdeabs{\Phi}{c}{\qity}{\memD_1}{\ity}{\memD_2}$ then (i) $\tdeabs{\Phi}{c}{\qity}{\memD_2}{\ity}{\memD_2}$ and (ii) $\tdeabs{\Phi}{c}{\qity}{\emptyset}{\ity}{\memD_2}$.
+\end{lemma}
+\begin{proof} Rule induction on (e).
+
+The proof follows by a straightforward application of the IH and re-application of the same rule in all cases except the following.
+
+\pfcase{(abs-ext-not-delegated-new), (abs-other-not-delegated-new)} In these cases, we can apply (abs-ty-stored) to arrive at (i). Conclusion (ii) follows by the IH.
+
+\pfcase{(abs-ty-stored)} In this case, conclusion (i) follows by re-applying (abs-ty-stored).
+Conclusion (ii) follows by applying Lemma \ref{lemma:type-translation-store-validitiy}, then inverting (ty-trans) and applying the IH.
+\end{proof}
+
+\begin{lemma}[Selective Type Translation Abstraction]
+\label{lemma:selective-type-translation-abstraction}
+If (a) $\sofkVX{\sqity{\qity}}{\kity}$ and (b) $\sval{\sqity{\qity}}{\es;\Upsilon;\Phi}$ and (c) $\vdash \Phi$ and (d) $\tytsvX{\memD_1}$ and (e) $\tdeabs{\Phi}{c}{\qity}{\memD_1}{\ity}{\memD_2}$ then (i) $\tytsvX{\memD_2}$ and given $\memD_1 \leadsto \delta_1 : \Delta_1$, (ii) $\memD_2 \leadsto \delta_1\delta_2 : \Delta_1\Delta_2$ and (iii) $\emptyset \vdash \delta_1\delta_2 : \Delta_1\Delta_2$ and (iv) if $\qity=\srep{\st}$ then $\Delta_1\Delta_2 \vdash \tau$.
+\end{lemma}
+\begin{proof} Rule induction on (e).
+
+\pfcase{(abs-parr)} In this case, we have $\qity=\srep{\sty{\rightharpoonup}{(\st_1, \st_2)}}$ and $\ity=\iparr{\tau_1}{\tau_2}$ and \begin{enumerate}[(1)]
+\item $\tdeabs{\Phi}{c}{\srep{\st_1}}{\memD_1}{\tau_1}{\memD'}$
+\item $\tdeabs{\Phi}{c}{\srep{\st_2}}{\memD'}{\tau_2}{\memD_2}$
+\end{enumerate}
+
+By Lemma \ref{lemma:static-canonical-forms}.8 on (a) and (b), we have (3) $\istype{\sty{\rightharpoonup}{(\st_1, \st_2)}}{\Phi}$.
+By Lemma \ref{lemma:static-canonical-forms}.7 on (3), we have (4) $\istype{\st_1}{\Phi}$ and (5) $\istype{\st_2}{\Phi}$.
+
+By Definition \ref{def:types} on (4), we have (6) $\sofkVX{\st_1}{\kty}$ and (7) $\sval{\st_1}{\es;\Upsilon;\Phi}$.
+
+By Definition \ref{def:types} on (5), we have (8) $\sofkVX{\st_2}{\kty}$ and (9) $\sval{\st_2}{\es;\Upsilon;\Phi}$.
+
+By (k-qity-parr) on (6), we have (10) $\sofkVX{\sqity{\srep{\st_1}}}{\kity}$. By (n-qity-parr) on (7), we have (11) $\sval{\sqity{\srep{\st_1}}}{\es;\Upsilon;\Phi}$.
+
+By (k-qity-parr) on (8), we have (12) $\sofkVX{\sqity{\srep{\st_2}}}{\kity}$. By (n-qity-parr) on (9), we have (13) $\sval{\sqity{\srep{\st_2}}}{\es;\Upsilon;\Phi}$.
+
+By the IH on (10), (11), (c), (d) and (1), we have (14) $\tytsvX{\memD'}$ and (15) $\memD' \leadsto \delta_1\delta' : \Delta_1\Delta'$ and (16) $\emptyset \vdash \delta_1\delta' : \Delta_1\Delta'$ and (17) $\Delta_1\Delta \vdash \tau_1$.
+
+By the IH on (12), (13), (c), (14) and (2), we have (i) $\tytsvX{\memD_2}$ and (ii) $\memD_2 \leadsto \delta_1\delta'\delta_2 : \Delta_1\Delta'\Delta_2$ and (iii) $\emptyset \vdash \delta_1\delta'\delta_2 : \Delta_1\Delta'\Delta_2$ and (18) $\Delta_1\Delta'\Delta_2 \vdash \tau_2$.
+
+By Weakening on (17), we have that (19) $\Delta_1\Delta'\Delta_2 \vdash \tau_1$ and so by (18) and (19) and the usual rules of internal type formation, we have (iv) $\Delta_1\Delta'\Delta_2 \vdash \iparr{\tau_1}{\tau_2}$.
+
+\pfcase{(abs-ext-delegated)} In this case, we have $\qity=\srep{\sty{\tc}{\sttyidx}}$ and \begin{enumerate}[(1)]
+\item $\tcdef{\tc}{\psi}{\tcstruct{\stx{schema}}{\omega}} \in \Phi$
+\item $\stx{schema}(\stx{tyidx}) \Downarrow_{\cdot;\emptyset;\Phi} \sqity{\qity'}$
+\item $\tdeabs{\Phi}{\tc}{\qity'}{\memD_1}{\tau}{\memD_2}$
+\item $\memD_2 \leadsto \delta : \Delta$
+\item $\Delta \vdash \tau$
+\end{enumerate}
+
+By Lemma \ref{lemma:well-defined-contexts-have-well-formed-signatures} on (1), we have that there exists a prefix $\Phi'$ of $\Phi$ such that (6) $\vdash \Phi'$ and (7) $\sofkz{\emptyset}{\emptyset}{\Phi'}{\stx{schema}}{\ktyidx \rightarrow \kity}$.
+By Corollary \ref{lemma:stability-of-kinding-3} on (6), (7) and (c), we have (8) $\sofkz{\emptyset}{\emptyset}{\Phi}{\stx{schema}}{\ktyidx \rightarrow \kity}$.
+
+By Lemma \ref{lemma:static-canonical-forms}.8 on (a) and (b), we have (9) $\istype{\sty{\tc}{\sttyidx}}{\Phi}$.
+By Lemma \ref{lemma:types-are-argument-independent} on (9), we have (10) $\sofkVXz{\sttyidx}{\ktyidx}$ and (11) $\sval{\sttyidx}{\es;\Upsilon;\Phi}$.
+
+By (k-ap) on (8) and (10), we have (12) $\sofkVXz{\stx{schema}(\sttyidx)}{\kity}$.
+
+By (tctx-ok-emp), we have (13) $\vdash_\Phi \emptyset$.
+
+By Corollary \ref{lemma:static-normalization-preservation} on (2), (12), (c) and (13), we have (14) $\sofkVXz{\sqity{\qity'}}{\kity}$.
+
+By Definition \ref{definition:static-normalization} on (2), we have (15) $\sval{\sqity{\qity'}}{\es;\Upsilon;\Phi}$.
+
+Applying the IH to (14), (15), (c), (d) and (3), which is well-founded by the metric defined below, we have (i), (ii) and (iii). By (5), we have (iv).
+
+\pfcase{(abs-ext-not-delegated-new)} In this case, we have $\qity=\srep{\sty{\tc}{\sttyidx}}$ and $\memD_2=\memD', \sty{\tc}{\sttyidx} \leftrightarrow [\delta]\tau'/\alpha$  and \begin{enumerate}[(1)]
+    \item $c\neq\tc$
+    \item $\sty{\tc}{\sttyidx} \notin \text{dom}(\memD_1)$
+    \item $\tcdef{\tc}{\psi}{\tcstruct{\stx{schema}}{\omega}} \in \Phi$
+    \item $\stx{schema}(\stx{tyidx}) \Downarrow_{\cdot;\emptyset;\Phi} \sqity{\qity'}$
+    \item $\tdeabs{\Phi}{c}{\qity}{\memD_1}{\tau'}{\memD'}$
+    \item $\memD' \leadsto \delta : \Delta$
+    \item $\Delta \vdash \tau'$
+\end{enumerate}
+By Lemma \ref{lemma:well-defined-contexts-have-well-formed-signatures} on (1), we have that there exists a prefix $\Phi'$ of $\Phi$ such that (6) $\vdash \Phi'$ and (7) $\sofkz{\emptyset}{\emptyset}{\Phi'}{\stx{schema}}{\ktyidx \rightarrow \kity}$.
+By Corollary \ref{lemma:stability-of-kinding-3} on (6), (7) and (c), we have (8) $\sofkz{\emptyset}{\emptyset}{\Phi}{\stx{schema}}{\ktyidx \rightarrow \kity}$.
+
+By Lemma \ref{lemma:static-canonical-forms}.8 on (a) and (b), we have (9) $\istype{\sty{\tc}{\sttyidx}}{\Phi}$.
+By Lemma \ref{lemma:types-are-argument-independent} on (9), we have (10) $\sofkVXz{\sttyidx}{\ktyidx}$ and (11) $\sval{\sttyidx}{\es;\Upsilon;\Phi}$.
+
+By (k-ap) on (8) and (10), we have (12) $\sofkVXz{\stx{schema}(\sttyidx)}{\kity}$.
+
+By (tctx-ok-emp), we have (13) $\vdash_\Phi \emptyset$.
+
+By Corollary \ref{lemma:static-normalization-preservation} on (2), (12), (c) and (13), we have (14) $\sofkVXz{\sqity{\qity'}}{\kity}$.
+
+By Definition \ref{definition:static-normalization} on (2), we have (15) $\sval{\sqity{\qity'}}{\es;\Upsilon;\Phi}$.
+
+Applying the IH to (14), (15), (c), (d) and (3), which is well-founded by the metric defined below, we have (16) $\tytsvX{\memD'}$ and (17) $\memD' \leadsto \delta_1\delta' : \Delta_1\Delta'$ and (18) $\emptyset \vdash \delta_1\delta' : \Delta_1\Delta'$.
+
+Applying (ty-store-int-ext) to (6) we have (19) $\memD_2 \leadsto (\delta_1\delta', [\delta_1\delta']\tau/\alpha) : \Delta_1\Delta', \alpha$, i.e. (ii).  By (7) and Weakening, we have (iv).
+
+By definition,  $[\delta_1\delta', [\delta_1\delta']\tau'/\alpha]\alpha=[\delta_1\delta']\tau'$.
+
+By Lemma \ref{lemma:type-translation-stores-are-idempotent} on (a-e), we have (20) $\tytransX{\st}{[\delta_1\delta']\tau'}$.
+
+Applying (ty-trans) to (9), (20) and (19), we have (21) $\tytransX{\sty{\tc}{\sttyidx}}{[\delta_1\delta']\tau'}$.
+
+By Assumption \ref{assumption:static-type-variable-substitution} on (7) and (iii), we have (22) $\emptyset \vdash [\delta_1\delta']\tau'$.
+
+Applying (tyts-ok-ext) to (16), (21) and (19) and (22), we have (i). By (tysub-ext) on (18) and (22), we have (iii).
+
+\pfcase{(abs-other-delegated)} In this case, $\qity=\srep{\sty{\otherc{m}{\kappa}}{(\stx{data}, \sqity{\qity'})}}$ and \begin{enumerate}
+    \item $\tdeabs{\Phi}{\otherc{m}{\kappa}}{\qity'}{\memD_1}{\tau}{\memD_2}$.
+    \item $\memD_2 \leadsto \delta : \Delta$
+    \item $\Delta \vdash \tau$
+\end{enumerate}
+
+By Lemma \ref{lemma:static-canonical-forms} iterated on (a) and (b), we have (4) $\sofkVX{\sqity{\qity'}}{\kity}$ and (5) $\sval{\sqity{\qity'}}{\es;\Upsilon;\Phi}$.
+Apply the IH to (4), (5), (c), (d) and (1).
+
+\pfcase{(abs-other-not-delegated-new)} In this case, $\qity=\srep{\sty{\otherc{m}{\kappa}}{(\stx{data}, \sqity{\qity'})}}$ and $\tau=\alpha$ and $\memD_2 = \memD', \sty{\otherc{m}{\kappa}}{(\stx{data}, \sqity{\qity'})} \leftrightarrow [\delta]\tau'/\alpha$ and \begin{enumerate}
+    \item $c \neq \otherc{m}{\kappa}$
+    \item $\sty{\otherc{m}{\kappa}}{(\stx{data}, \sqity{\qity'})} \notin \text{dom}(\memD)$
+    \item $\tdeabs{\Phi}{c}{\qity'}{\memD_1}{\tau'}{\memD'}$.
+    \item $\memD' \leadsto \delta : \Delta$
+    \item $\Delta \vdash \tau'$
+\end{enumerate}
+
+By Lemma \ref{lemma:static-canonical-forms} iterated on (a) and (b), we have (6) $\istype{\sty{\otherc{m}{\kappa}}{(\stx{data}, \sqity{\qity'})}}{\Phi}$ and (7) $\sofkVX{\sqity{\qity'}}{\kity}$ and (8) $\sval{\sqity{\qity'}}{\es;\Upsilon;\Phi}$.
+Applying the IH to (7), (8), (c), (d) and (3), we have (9) $\tytsvX{\memD'}$ and (10) $\memD' \leadsto \delta_1\delta' : \Delta_1\Delta'$ and (11) $\emptyset \vdash \delta_1\delta' : \Delta_1\Delta'$.
+
+Applying (ty-store-int-ext) to (4) we have (12) $\memD_2 \leadsto (\delta_1\delta', [\delta_1\delta']\tau/\alpha) : \Delta_1\Delta', \alpha$, i.e. (ii). By (5) and Weakening, we have (iv).
+
+By definition, (13) $[\delta_1\delta', [\delta_1\delta']\tau'/\alpha]\alpha=[\delta_1\delta']\tau'$.
+
+Applying Lemma \ref{lemma:type-translation-stores-are-idempotent} on (a-e), we have its conclusion as (14).
+
+Applying (ty-trans) to (6), (14) and (12), we have (15) $\tytransX{\sty{\otherc{m}{\kappa}}{(\stx{data}, \sqity{\qity'})}}{[\delta]\tau'}$.
+
+By Assumption \ref{assumption:static-type-variable-substitution} on (5) and (iii), we have (16) $\emptyset \vdash [\delta_1\delta']\tau'$.
+
+Applying (tyts-ok-ext) to (6), (1), (15) and (16), we have (i). By (tysub-ext) on (11) and (16), we have (iii).
+
+\pfcase{(abs-ty-stored)} Apply \ref{lemma:type-translation-store-validitiy} to (d).
+
+\pfcases{(abs-i-alpha), (abs-i-t), (abs-i-unit)} Apply (d).
+
+\pfcases{(abs-i-parr), (abs-i-forall), (abs-i-mu), (abs-i-prod), (abs-i-sum)} These cases follow by iterated application of the IH and Lemma \ref{lemma:static-canonical-forms}.8.
+\end{proof}
+
+The induction in this case is justified for cases (-not-delegated-new) because the number of sub-terms of the form $\sty{c}{\st}$ is strictly decreasing by kinding.
+
+\subsection{Translation Validation}
+\subsubsection{Type and Typing Context Translation Validation}
+\begin{lemma}[Type Translation]
+\label{lemma:type-translation}
+If (a) $\vdash \Phi$ and (b) $\tytransX{\st}{\tau}$ then $\emptyset \vdash \tau$.
+\end{lemma}
+\begin{proof} Rule induction on (b). There is only one rule, so we have $\tau=[\delta]\tau'$ \begin{enumerate}[(1)]
+\item $\istype{\st}{\Phi}$
+\item $\tytsvX{\memD}$
+\item $\tdeabs{\Phi}{c}{\srep{\st}}{\memD}{\tau'}{\memD'}$
+\item $\memD' \leadsto \delta : \Delta$
+\end{enumerate}
+Taking $\memD=\emptyset$ and applying Lemma \ref{lemma:selective-type-translation-abstraction}, we have $\Delta \vdash \tau'$ and $\emptyset \vdash \delta : \Delta$.
+Applying Assumption \ref{lemma:internal-type-substitution-on-types}, we have our conclusion.
+\end{proof}
+
+\begin{lemma}[Typing Context Translation]
+If (a) $\vdash \Phi$ and (b) $\ctxtransX{\Upsilon}{\Gamma}$ then $\emptyset \vdash \Gamma$.\end{lemma}
+\begin{proof} Rule induction on the assumption.
+
+\pfcase{(ctx-trans-emp)} We have $\emptyset \vdash \Gamma$ by (ictx-emp).
+
+\pfcase{(ctx-trans-ext)} We have that $\Upsilon=\Upsilon', x : \st$ and $\Gamma=\Gamma', x : \tau$ and (1) $\ctxtransX{\Upsilon'}{\Gamma}$ and (2) $\tytransX{\st}{\tau}$.
+
+Applying the IH to (a) and (1), we have (3) $\emptyset \vdash \Gamma'$.
+
+Applying Lemma \ref{lemma:type-translation} to (a) and (2) we have (4) $\emptyset \vdash \tau$. Applying (ictx-ext) to (3) and (4), we have $\emptyset \vdash \Gamma', x : \tau$.
+\end{proof}
+
+\begin{lemma}[Typing Context Translation Membership]
+\label{lemma:typing-context-translation-membership}
+If (a) $\vdash \Phi$ and (b) $\ctxtransX{\Upsilon}{\Gamma}$ and (c) $x : \sty{c}{\st} \in \Upsilon$ then (i) $\tytransX{\st}{\tau}$ and (ii) $x : \tau \in \Gamma$.
+\end{lemma}
+\begin{proof} Rule induction on (b) and (c).
+
+\pfcase{(ctx-trans-emp)} Does not apply by (c).
+
+\pfcase{(ctx-trans-ext)} We have that $\Upsilon=\Upsilon', x' : \st'$ and  (1) $\ctxtransX{\Upsilon'}{\Gamma'}$. We have two cases by (c):
+
+\begin{quote}
+    \pfcase{$x=x'$} We have our conclusion as the second premise.
+
+    \pfcase{$x\neq x'$ and (2) $x : \sty{c}{\st} \in \Upsilon'$} Applying the IH to (1) and (2), we have our conclusion.
+\end{quote}
+\end{proof}
+
+\subsubsection{Selectively Abstracted Term Translations}
+The rules for selectively abstracted term translations are specified in Figures \ref{fig:selectively-abstracted-term-translations-1} and \ref{fig:selectively-abstracted-term-translations-2}.
+The auxiliary judgement for validating a type and term translation store together is $\tmtsvX{\memD}{\memG}$, specified in Figure \ref{fig:term-translation-store-validity}.
+The internalization judgement $\memG \leadsto \gamma : \Gamma$ is specified in Figure \ref{fig:term-translation-store-Internalization}.
+
+\begin{lemma}
+\label{lemma:get-D}
+If $\tmtsvX{\memD}{\memG}$ then $\tytsvX{\memD}$.
+\end{lemma}
+\begin{proof}
+    The conclusion is a premise of each rule for the assumption.
+\end{proof}
+
+\begin{lemma}[Argument Translation]
+\label{lemma:argument-translation}
+If (a) $\tmtsvX{\memD}{\memG}$ and (b) $n : \st \leadsto \iota/x : \tau$ then \begin{enumerate}[(i)]
+\item $\tytsvX{\memD}$
+\item $\tdeabs{\Phi}{c}{\srep{\st}}{\memD}{\tau}{\memD}$
+\item $\memD \leadsto \delta : \Delta$
+\item $\tytransX{\st}{[\delta]\tau}$
+\item $\keyw{nth}[n](\es)=e$
+\item $\eanaX{e}{\st}{\iota}$
+\end{enumerate}
+\end{lemma}
+\begin{proof} Rule induction on (a) and (b).
+
+\pfcase{(tmts-ok-emp)} Does not apply by (b).
+
+\pfcase{(tmts-ok-ext)} We have $\memG=\memG', n' : \st' \leadsto \iota'/x' :\tau'$ There are two cases by (b).
+\begin{quote}
+    \pfcase{$n=n'$} The conclusions are the premises.
+
+    \pfcase{$n\neq n'$} Apply the IH.
+\end{quote}
+\end{proof}
+
+\begin{lemma}[Selectively Abstracted Term Translation]
+\label{lemma:selectively-abstracted-term-translations}
+If (a) $\sofkVX{\sqitm{\qitm}}{\kitm}$ and (b) $\sval{\sqitm{\qitm}}{\es;\Upsilon;\Phi}$ and (c) $|\es|=n$ and (d) $\vdash \Phi$ and (e) $\ctxtransX{\Upsilon}{\Gamma}$ and (f) $\tmtsvX{\memD}{\memG}$ and (g) $\edeabs{c}{\es;\Upsilon;\Phi}{\qitm}{\memD}{\memG}{\itm}{\memD'}{\memG'}$ then $\tmtsvX{\memD'}{\memG'}$.
+\end{lemma}
+\begin{proof} Rule induction on (g).
+
+\pfcase{(abs-anatrans-new)} In this case, $\qitm=\anatrans{n'}{\st}$ and $\itm=x$ and $\memG'=\memG, n' : \st \leadsto \iota/x : \tau$ and \begin{enumerate}
+    \item $n' \notin \text{dom}(\memG)$
+    \item $\keyw{nth}[n'](\es)=e$
+    \item $\eanaX{e}{\st}{\iota}$
+    \item $\tdeabs{\Phi}{c}{\srep{\st}}{\memD}{\tau}{\memD'}$
+\end{enumerate}
+
+By Lemma \ref{lemma:get-D} on (f), we have (5) $\tytsvX{\memD}$.
+
+By iterated application of Lemma \ref{lemma:static-canonical-forms}, we have (6) $\sofkVX{\sqity{\srep{\st}}}{\kity}$ and (7) $\sval{\sqity{\srep{\st}}}{\es;\Upsilon;\Phi}$ and (8) $\istype{\st}{\Phi}$.
+
+By Lemma \ref{lemma:selective-type-translation-abstraction} on (6), (7), (d), (5) and (4) we have (9) $\tytsvX{\memD'}$ and (10) $\memD' \leadsto \delta : \Delta$.
+
+By Lemma \ref{lemma:type-translation-stores-are-idempotent} on (6), (7), (d) and (4) we have (11) $\tdeabs{\Phi}{c}{\srep{\st}}{\emptyset}{\tau}{\memD'}$ and (12) $\tdeabs{\Phi}{c}{\srep{\st}}{\memD'}{\tau}{\memD'}$.
+
+By (ty-trans) on (8), (11) and (10), we have (13) $\tytransX{\st}{[\delta]\tau}$.
+
+Apply (tmts-ok-ext) to (5), (7), (10), (13), (2) and (3).
+
+\pfcase{(abs-anatrans-stored)} We have $\memD'=\memD$ and $\memG'=\memG$. Apply (e).
+
+\pfcase{(abs-syntrans-new)}  In this case, $\qitm=\syntrans{n'}$ and $\itm=x$ and $\memG'=\memG, n' : \st \leadsto \iota/x : \tau$ and \begin{enumerate}
+    \item $n' \notin \text{dom}(\memG)$
+    \item $\keyw{nth}[n'](\es)=e$
+    \item $\esynX{e}{\st}{\iota}$
+    \item $\tdeabs{\Phi}{c}{\srep{\st}}{\memD}{\tau}{\memD'}$
+\end{enumerate}
+
+By Lemma \ref{lemma:get-D} on (f), we have (5) $\tytsvX{\memD}$.
+
+By Lemma \ref{lemma:type-synthesis} and kinding rules, we have (6) $\sofkVX{\sqity{\srep{\st}}}{\kity}$ and (7) $\sval{\sqity{\srep{\st}}}{\es;\Upsilon;\Phi}$ and (8) $\istype{\st}{\Phi}$.
+
+By Lemma \ref{lemma:selective-type-translation-abstraction} on (6), (7), (d), (5) and (4) we have (9) $\tytsvX{\memD'}$ and (10) $\memD' \leadsto \delta : \Delta$.
+
+By Lemma \ref{lemma:type-translation-stores-are-idempotent} on (6), (7), (d) and (4) we have (11) $\tdeabs{\Phi}{c}{\srep{\st}}{\emptyset}{\tau}{\memD'}$ and (12) $\tdeabs{\Phi}{c}{\srep{\st}}{\memD'}{\tau}{\memD'}$.
+
+By (ty-trans) on (8), (11) and (10), we have (13) $\tytransX{\st}{[\delta]\tau}$.
+
+Apply (tmts-ok-ext) to (5), (7), (10), (13), (2) and (3).
+
+\pfcase{(abs-syntrans-stored)} We have $\memD'=\memD$ and $\memG'=\memG$. Apply (e).
+
+\pfcases{(abs-x), (abs-triv)} We have $\memD'=\memD$ and $\memG'=\memG$. Apply (e).
+
+\pfcases{(all remaining cases)} These cases follow by iterative application of the IH and Lemma \ref{lemma:static-canonical-forms}.9.
+
+\end{proof}
+
+\subsubsection{Translation Validation}
+The translation validation judgement is specified in Figure \ref{fig:translation-validation} and discussed together with type-preserving translation below.
+
+\subsection{Type-Preserving Translation and Type Safety}
+\begin{theorem}[Type-Preserving Translation] 
+\label{thm:type-preserving-translation}
+\begin{enumerate}
+\item[] ~
+\item[1)] If (a) $\vdash \Phi$ and (b) $\ctxtransX{\Upsilon}{\Gamma}$ and (c) $\esynX{e}{\st}{\iota}$ then (i) $\tytransX{\st}{\tau}$ and (ii) $\emptyset~\Gamma \vdash \iota : \tau$.
+\item[2)] If (a) $\vdash \Phi$ and (b) $\ctxtransX{\Upsilon}{\Gamma}$ and (c) $\istype{\st}{\Phi}$ and (d) $\eanaX{e}{\st}{\iota}$ then (i) $\tytransX{\st}{\tau}$ and (ii) $\emptyset~\Gamma \vdash \iota : \tau$.
+\end{enumerate}
+\end{theorem}
+\begin{proof}
+\pfcase{(subsume)} In this case of 2), we have that (1) $\esynX{e}{\st}{\iota}$. Apply the IH to (a), (b) and (1).
+
+\pfcase{(ascribe)} In this case of 1), we have that $e=e' : \st'$ and (1) $\sofkVXz{\st'}{\kty}$ and (2) $\st' \Downarrow_{\cdot;\emptyset;\Phi} \st$ and (3) $\eanaX{e}{\st}{\iota}$.
+
+Applying Definition \ref{definition:static-normalization} to (2), we have (4) $\sval{\st}{\cdot;\emptyset;\Phi}$.
+Applying Corollary \ref{lemma:static-normalization-preservation} to (1) and (2), we have (5) $\sofkVXz{\st}{\kty}$.
+Applying Definition \ref{def:types} to (4) and (5), we have (6) $\istype{\st}{\Phi}$.
+Apply the IH to (a), (b), (6) and (3).
+
+\pfcase{(syn-var)} In this case of 1), We have $e=x$ and $\iota=x$ and (1) $x : \st \in \Upsilon$.
+Applying Lemma \ref{lemma:typing-context-translation-membership} to (a) and (b) and (1), we have (i) and (2) $x : \tau \in \Gamma$.
+Applying the standard typing variable rule for the IL to (2), we have $\emptyset \Gamma \vdash x : \tau$.
+
+\pfcase{(ana-fix)} In this case of 2), we have $e=\efix{x}{e'}$ and $\iota=\ifix{\tau}{x}{\iota'}$ and (1) $\eana{\Upsilon, x : \st}{e'}{\st}{\iota'}$ and (2) $\tytransX{\st}{\tau}$, i.e. (i).
+
+Applying (ctx-trans-ext) to (b) and (2), we have (3) $\ctxtransX{\Upsilon, x : \st}{\Gamma, x : \tau}$.
+
+Applying Lemma \ref{lemma:types-with-translations-are-types} to (2), we have (4) $\istype{\st}{\Phi}$.
+
+Applying the IH to (a), (3), (4) and (1), we have (5) $\emptyset~\Gamma, x : \tau \vdash \iota' : \tau$.
+
+Applying Lemma \ref{lemma:type-translation} to (2), we have (6) $\emptyset \vdash \tau$.
+
+By the standard typing rule for fixpoints in the IL on (6) and (5), we have (i) $\emptyset~\Gamma, x : \tau \vdash \ifix{\tau}{x}{\iota'}$.
+
+\pfcase{(syn-lam)} In this case of 1), we have $e=\elam{\st_1}{x}{e'}$ and $\st=\sty{\rightharpoonup}{(\st_1, \st_2)}$ and $\iota=\ilam{\tau_1}{x}{\iota'}$ and (1) $\sofkVXz{\st_1}{\kty}$ and (2) $\st_1 \Downarrow_{\cdot;\emptyset;\Phi} \st_1'$ and (3) $\esyn{\Upsilon, x : \st_1'}{e'}{\st_2}{\iota'}$ and (4) $\tytransX{\st_1'}{\tau_1}$.
+
+Applying Definition \ref{definition:static-normalization} to (2), we have (5) $\sval{\st_1'}{\cdot;\emptyset;\Phi}$.
+Applying Corollary \ref{lemma:static-normalization-preservation} to (1) and (2), we have (6) $\sofkVXz{\st_1'}{\kty}$.
+Applying Definition \ref{def:types} to (5) and (6), we have (7) $\istype{\st_1'}{\Phi}$.
+
+Applying (ctx-trans-ext) to (b) and (4), we have (8) $\ctxtransX{\Upsilon, x : \st_1'}{\Gamma, x : \tau_1}$.
+
+Applying the IH to (a), (8) and (3), we have (9) $\tytransX{\st_2}{\tau_2}$ and (10) $\emptyset~\Gamma, x : \tau_1 \vdash \iota' : \tau_2$.
+
+By Lemma \ref{fig:type-translation} on (9), we have (11) $\emptyset \vdash \tau_2$.
+
+By the standard typing rule for lambdas in the IL on (10) and (9), we have (ii).
+
+%\pfcase{(ana-lam)} This case follows the script above.\todo{copy pasta}
+
+\pfcase{(syn-ap)} In this case of 1), we have $e=e_1(e_2)$ and $\iota=\iota_1(\iota_2)$ and (1) $\esynX{e_1}{\sty{\rightharpoonup}{(\st_1, \st_2)}}{\iota_1}$ and (2) $\eanaX{e_2}{\st_1}{\iota_2}$.
+
+Applying the IH to (a), (b) and (1), we have (3) $\tytransX{\st_1}{\tau_1}$ and (4) $\emptyset~\Gamma \vdash \iota_1 : \tau_1$.
+
+Applying the IH to (a), (b) and (2), we have (5) $\tytransX{\st_2}{\tau_2}$ and (6) $\emptyset~\Gamma \vdash \iota_2 : \tau_2$.
+
+By the standard typing rule for application in the IL on (4) and (6), we have (ii).
+
+\pfcase{(ana-intro)} In this case of 2), we have $e=\eintro{\sttmidx}{\es}$ and $\st=\sty{\tc}{\sttyidx}$ and \begin{enumerate}[(1)]
+  \item $\tcdef{\tc}{\tcsig{\ktyidx}{\chi}}{\tcstruct{\stx{schema}}{\omega}} \in \Phi$
+  \item $\introsig{\klitidx} \in \chi$
+  \item $\sofkz{\emptyset}{\emptyset}{\Phi}{\stmidx}{\klitidx}$
+  \item $\keyw{ana~intro}={\stx{def}} \in \omega$
+  \item $|\es| = n$
+  \item $\keyw{args}(n)=\stx{args}$
+  \item $\stx{def}(\sttyidx)(\stmidx)(\stx{args}) \Downarrow_{\es;\Upsilon;\Phi} \sqitm{\qitm}$
+  \item $\trvalidate{\es;\Upsilon;\Phi}{\tc}{\qitm}{{\sty{\tc}{\sttyidx}}}{\iota}$
+\end{enumerate}
+
+Apply Lemma \ref{lemma:tycon-context-validity} to (1), we have that there exists a prefix $\Phi'$ of $\Phi$ such that (9) $\vdash_{\Phi', \tcdef{\tc}{\tcsig{\ktyidx}{\chi}}{\tcstruct{\stx{schema}}{\omega}}} \omega \sim \psi$ and (10) $\vdash \Phi'$.
+
+Apply Lemma \ref{lemma:stability-of-opcon-structure-validity-3} to (10), (9) and (a), we have (11) $\vdash_{\Phi} \omega \sim \psi$
+
+Applying Lemma \ref{lemma:intro-opcon-existence-and-validity} to (11) and (4), we have (12) $\sofkVXz{\stx{def}}{\ktyidx \rightarrow \klitidx \rightarrow \kargs \rightarrow \kitm}$.
+
+Applying Lemma \ref{lemma:weakening-of-argument-bounds} to (12), we have (13) $\sofkVXz{\stx{def}}{\ktyidx \rightarrow \klitidx \rightarrow \kargs \rightarrow \kitm}$.
+
+Applying Lemma \ref{lemma:static-canonical-forms} on (c), we have that (14) $\sofkVX{\sttyidx}{\ktyidx}$.
+
+Apply Lemma \ref{lemma:argument-interface-lists}, we have that (15) $\sofkVX{\stx{args}}{\kargs}$.
+
+Applying (k-ap) three times with (13), (14), (3) and (15), we have that (16) $\sofkVX{\stx{def}(\sttyidx)(\stmidx)(\stx{args})}{\kitm}$.
+
+Applying Lemma \ref{lemma:static-normalization-preservation} to (16) and (7), we have (17) $\sofkVX{\sqitm{\qitm}}{\kitm}$.
+
+Applying Definition \ref{definition:static-normalization} to (7), we have (18) $\sval{\sqitm{\qitm}}{\es;\Upsilon;\Phi}$.
+
+Apply Lemma \ref{lemma:translation-validation} to (a), (b), (c), (17), (18), (5) and (7).
+
+\pfcase{(syn-targ)} In this case of 1), we have $e=\etarg{\opname{op}}{\stx{tmidx}}{e_\text{targ}}{\es}$ and \begin{enumerate}[(1)]
+\item $\esynX{e_\text{targ}}{\sty{\tc}{\sttyidx}}{\iota_\text{targ}}$ and 
+\item $\tcdef{\tc}{\tcsig{\ktyidx}{\chi}}{\tcstruct{\stx{schema}}{\omega}} \in \Phi$ and
+\item $\opsig{\opname{op}}{\klitidx} \in \chi$ and
+\item $\sofkz{\emptyset}{\emptyset}{\Phi}{\stmidx}{\klitidx}$ and
+\item $\keyw{syn~}\opname{op}={\stx{def}} \in \omega$ and
+\item $|e_\text{targ}; \es| = n$ and
+\item $\keyw{args}(n)=\stx{args}$ and
+\item $\stx{def}(\sttyidx)(\stmidx)(\stx{args}) \Downarrow_{(e_\text{targ}; \es);\Upsilon;\Phi} (\st, \sqitm{\qitm})$.
+\end{enumerate}
+
+Applying Lemma \ref{lemma:type-synthesis} to (a), (b) and (1), we have (9) $\istype{\sty{\tc}{\sttyidx}}{\Phi}$.
+By Lemma \ref{lemma:static-canonical-forms}.7 on (9), we have (10) $\sofkVXz{\sttyidx}{\ktyidx}$.
+Applying Lemma \ref{lemma:weakening-of-argument-bounds} to (10), we have (11) $\sofkVX{\sttyidx}{\ktyidx}$.
+
+Applying Lemma \ref{lemma:tycon-context-validity} to (2), we have that there is a prefix $\Phi'$ of $\Phi$ such that (12) $\vdash \Phi'$ and (13) $\vdash \Phi', \tcdef{\tc}{\tcsig{\ktyidx}{\chi}}{\tcstruct{\stx{schema}}{\omega}}$ and (14) $\vdash_{\Phi', \tcdef{\tc}{\tcsig{\ktyidx}{\chi}}{\tcstruct{\stx{schema}}{\omega}}} \omega \sim \tcsig{\ktyidx}{\chi}$.
+
+Applying Lemma \ref{lemma:well-defined-contexts-have-well-formed-signatures} to (12), we have (15) $\tccsigok{\Phi', \tcdef{\tc}{\tcsig{\ktyidx}{\chi}}{\tcstruct{\stx{schema}}{\omega}}}$.
+Applying Lemma \ref{lemma:well-defined-contexts-have-well-formed-signatures} to (a), taking by the definition of prefixing that $\Phi=(\Phi', \tcdef{\tc}{\tcsig{\ktyidx}{\chi}}{\tcstruct{\stx{schema}}{\omega}})\Phi''$, we have (16) $\tccsigok{(\Phi', \tcdef{\tc}{\tcsig{\ktyidx}{\chi}}{\tcstruct{\stx{schema}}{\omega}})\Phi''}$.
+Applying Lemma \ref{lemma:stability-of-opcon-structure-validity} to (15), (14) and (16), we have that (17) $\vdash_{\Phi} \omega \sim \tcsig{\ktyidx}{\chi}$.
+
+Applying Lemma \ref{lemma:targeted-opcon-validity} to (17) and (5), we have that (18) $\sofkz{\emptyset}{\emptyset}{\Phi}{\stx{def}}{\ktyidx \rightarrow \klitidx \rightarrow \kargs \rightarrow (\kty \times \kitm)}$.
+Applying Lemma \ref{lemma:weakening-of-argument-bounds} to (18), we have (19) $\sofk{\emptyset}{\emptyset}{\Phi}{\stx{def}}{\ktyidx \rightarrow \klitidx \rightarrow \kargs \rightarrow (\kty \times \kitm)}$.
+
+Applying Lemma \ref{lemma:argument-interface-lists} to (7), we have that (20) $\sofkVX{\stx{args}}{\kargs}$.
+
+Applying (k-ap) three times with (19), (11), (4) and (20), we have that (21) $\sofkVX{\stx{def}(\sttyidx)(\stx{tmidx})(\stx{args})}{\kty\times\kitm}$.
+
+Applying Corollary \ref{lemma:static-normalization-preservation} to (8), (21), (a) and (b), we have that (22) $\sofkVX{(\st, \sqitm{\qitm})}{\kty\times\kitm}$.
+By Definition \ref{definition:static-normalization} on (8), we have (23) $\sval{(\st, \sqitm{\qitm})}{(e_\text{targ}; \es); \Upsilon; \Phi}$.
+
+Applying Lemma \ref{lemma:static-canonical-forms} to (23) and (22), we have that (24) $\sofkVX{\st}{\kty}$ and (25) $\sval{\st}{(e_\text{targ}; \es); \Upsilon; \Phi}$ and (26) $\sofkVX{\sqitm{\qitm}}{\kitm}$ and (27) $\sval{\sqitm{\qitm}}{(e_\text{targ}; \es);\Upsilon;\Phi}$.
+Applying Definition \ref{def:types} to (24) and (25) we have (28) $\istype{\st}{\Phi}$.
+
+Apply Lemma \ref{lemma:translation-validation} to (a), (b), (28), (26), (27), (6) and (8).
+
+\pfcase{(ana-intro-other)} In this case, we have that $e=\eintro{\sqitm{\qitm}}{\es}$ and $\st=\sty{\otherc{m}{\kappa}}{\sttyidx}$ and \begin{enumerate}[(1)]
+    \item $|\es|=n$
+    \item $\sofkVX{\sqitm{\qitm}}{\kitm}$
+    \item $\sval{\sqitm{\qitm}}{\es; \Upsilon; \Phi}$
+    \item $\trvalidate{\es;\Upsilon;\Phi}{\keyw{other}[m;\kappa]}{\qitm}{{\sty{\keyw{other}[m;\kappa]}{\sttyidx}}}{\iota}$
+\end{enumerate}
+
+Apply Lemma \ref{lemma:translation-validation} to (a), (b), (c), (2), (3), (1) and (4).
+
+\pfcase{(syn-targ-other)} In this case, we have that $e=\etarg{\opname{op}}{\sqitm{\qitm}}{e_\text{targ}}{\es}$ and \begin{enumerate}[(1)]
+    \item $\esynX{e_\text{targ}}{\sty{\keyw{other}[m;\kappa]}{\sttyidx}}{\iota_\text{targ}}$
+    \item $|e_\text{targ}; \es| = n$
+    \item $\sofk{\emptyset}{\emptyset}{\Phi}{\sqitm{\qitm}}{\kitm}$
+    \item $\sval{\sqitm{\qitm}}{(e_\text{targ}; \es); \Upsilon; \Phi}$
+    \item $\istype{\st}{\Phi}$
+    \item $\trvalidate{(e_\text{targ}; \es);\Upsilon;\Phi}{\keyw{other}[m;\kappa]}{\qitm}{{\st}}{\iota}$
+\end{enumerate}
+
+Apply Lemma \ref{lemma:translation-validation} to (a), (b), (9), (3), (4), (2), (5) and (6).
+\end{proof}
+
+\begin{lemma}[Translation Validation]
+\label{lemma:translation-validation}
+If (a) $\vdash \Phi$ and (b) $\ctxtransX{\Upsilon}{\Gamma}$ and (c) $\istype{\st}{\Phi}$ and (d) $\sofkVX{\sqitm{\qitm}}{\kitm}$ and (e) $\sval{\sqitm{\qitm}}{\es;\Upsilon;\Phi}$ and (f) $|\es|=n$ and (g) $\vdash_{\es;\Upsilon;\Phi}^c \qitm : \st \leadsto \iota$ then (i) $\tytransX{\st}{\tau}$ and (ii) $\emptyset~\Gamma \vdash \iota : \tau$.
+\end{lemma}
+\begin{proof}
+    Rule induction on (g). There is one rule, so we have \begin{enumerate}[(1)]
+        \item $\tdeabs{\Phi}{c}{\srep{\st}}{\emptyset}{\ity_\text{abs}}{\memD}$
+        \item $\edeabs{c}{{\es;\Upsilon;\Phi}}{\qitm}{\memD}{\emptyset}{\iota_\text{abs}}{\memD'}{\memG}$
+        \item $\memD' \leadsto \delta : \Delta_\text{abs}$
+        \item $\memG \leadsto \gamma : \Gamma_\text{abs}$
+        \item $\Delta_\text{abs}~\Gamma_\text{abs} \vdash \iota_\text{abs} : \tau_\text{abs}$
+    \end{enumerate}
+
+    Applying Lemma \ref{lemma:selective-type-translation-abstraction} on (1), we have that (6) $\tytsvX{\memD}$.
+
+    Applying Lemma \ref{lemma:selectively-abstracted-term-translations} to (6), we have (7) $\tmtsvX{\memD}{\memG}$.
+
+    Applying Lemma \ref{lemma:BINGO}, we have that (8) $\memD \leadsto \delta : \Delta_\text{abs}$ and (9) $\memG \leadsto \gamma : \Gamma_\text{abs}$ and (10) $\emptyset \vdash \delta : \Delta_\text{abs}$ and (11) $\Delta~\Gamma \vdash \gamma : \Gamma_\text{abs}$.
+
+    By Assumptions \ref{assumption:internal-term-substituions} and \ref{lemma:internal-type-substitutions-on-terms} and (5) and (10) and (11), we have that $\emptyset~\Gamma \vdash [\delta][\gamma]\iota_\text{abs} : \tau_\text{abs}$.
+\end{proof}
+
+\begin{lemma}
+\label{lemma:BINGO}
+    If (a) $\tmtsvX{\memD}{\memG}$ and (b) $\ctxtransX{\Upsilon}{\Gamma}$ and (c) $\memD \leadsto \delta : \Delta_\text{abs}$ and (d) $\memG \leadsto \gamma : \Gamma_\text{abs}$ then (i) $\emptyset \vdash \delta : \Delta$ and (ii) $\Delta~\Gamma \vdash \gamma : \Gamma_\text{abs}$.
+\end{lemma}\begin{proof} Inversion on (a).
+
+    (i) follows from Lemma \ref{lemma:type-translation-store-validitiy}.
+    (ii) follows by inductive appeal to Theorem \ref{thm:type-preserving-translation}, justified by the same metric as Theorem \ref{lemma:type-synthesis}.
+\end{proof}
+
+% \subsection{Stability}
+% \todo{}
+
+% \subsection{Unicity}
+% Need determinism of static dynamics for this.
+
+% The rules are structured so that if a term is well-typed, both its type and translation are unique.%\todo{could move this whole thing to supplement if room needed}
+% \begin{theorem}[Unicity]
+% If $\vdash \Phi$ and $\Upsilon~\mathtt{ctx}_{\Phi} \leadsto \Gamma$ and $\vdash_\Phi \st \leadsto \tau$ and $\vdash_\Phi \st' \leadsto \tau'$ and $\eanaX{e}{\st}{\iota}$ and $\eanaX{e}{\st'}{\iota'}$ then $\st = \st'$ and $\tau=\tau'$ and $\iota = \iota'$.
+% \end{theorem}
+
+\subsection{Conservativity}
+\subsubsection{Backprojection}
+To avoid having to give a large number of rules that do nothing interesting, we define backprojection on static terms definitionally.
+In the discussion below, let $\Phi'=\Phi, \tcdef{\tc'}{\tcsig{\ktyidx}{\chi}}{\tcstruct{\stx{schema}}{\omega}}}{\st}{\kappa}$.
+
+\newcommand{\bprj}[4]{{}^{m}\mathcal{B}_{#1}^{#2}(#3)=#4}
+\newcommand{\bprjX}[2]{\bprj{\Phi'}{\tc'}{#1}{#2}}
+\begin{definition}[Backprojection (Static values)]
+We write $\bprjX{\st}{\st'}$ for the backprojection of $\st$ when $\sofk{\emptyset}{\emptyset}{\Phi'}{\st}{\kappa}$ and $\sval{\st}{\argEnv}$ and $\stx{schema}(\sttyidx) \Downarrow_{\cdot;\emptyset;\Phi'} \sqity{\qity}$.
+We generate $\st'$ by replacing all sub-terms of the form $\sty{\tc'}{\sttyidx}$ with a sub-term of the form $\sty{\otherc{m}{\ktyidx}}{(\sttyidx, \sqity{\qity})}$.
+\end{definition}
+
+All interesting properties about static values (in particular, that they are values and their kind) are preserved under backprojection.
+
+\begin{lemma}
+If $\sofk{\emptyset}{\emptyset}{\Phi'}{\st}{\kappa}$ and $\sval{\st}{\argEnv}$ and $\bprjX{\st}{\st'}$ then $\sofkVX{\st'}{\kappa}$ and $\sval{\st'}{\argEnv}$.
+\end{lemma}
+
+For types, backprojection preserves the type translation (essentially trivially, since the translational internal type is contained directly in the index) and for all types other than those constructed by $\tc$, their tycons.
+
+\begin{lemma}
+If $\sofk{\emptyset}{\emptyset}{\Phi'}{\st}{\kty}$ and $\sval{\st}{\argEnv}$ and $\tytrans{\Phi'}{\st}{\tau}$ and $\bprjX{\st}{\st'}$ then $\tytransX{\st}{\tau}$ and if $\st=\sty{c}{\sttyidx}$ and $c\neq\tc'$ then $\bprjX{\sttyidx}{\sttyidx'}$ and $\st'=\sty{c}{\sttyidx'}$.
+\end{lemma}
+
+We can map backprojection over typing contexts in the same way.
+\begin{definition}[Backprojection (Typing Contexts)]
+We write $\bprjX{\Upsilon}{\Upsilon'}$ when $\vdash_\Phi \Upsilon$ for the mapping of the backprojection operation over all types in $\Upsilon$.
+\end{definition}
+
+\begin{lemma} 
+If $\ctxtrans{\Phi'}{\Upsilon}{\Gamma}$ and $\bprjX{\Upsilon}{\Upsilon'}$ then $\ctxtrans{\Phi}{\Upsilon}{\Gamma}$.
+\end{lemma}
+
+\begin{definition}[Modular Propositions]
+We say a proposition $P(\Gamma, \st, \iota)$ is \emph{modular} under $\Phi$ if for all $\Phi''$ such that $\vdash \Phi\Phi''$ and all $\tc \in \text{dom}(\Phi'')$, we have that if $\bprj{\Phi\Phi''}{\tc'}{\st}{\st'}$ and $P(\Gamma, \st', \iota)$ then $P(\Gamma, \st, \iota)$.
+\end{definition}
+
+Note that for propositions constrained to only hold for $\st : \kappa$ such that $\kappa$ does not contain $\kty$, the proposition is modular trivially because backprojection is the identity relation in that case.
+In other cases, the proposition simply must not ``know'' about future tycons or treat ``other'' tycons in a special way (e.g.
+counting how many there are).
+We expect that this will generally be straightforward to establish.
+Note that this is a property of the invariant, and can thus be established modularly, not a proof obligation for every composition.
+
+\begin{definition}[Backprojection (External Terms)]
+If $\vdash \Phi'$ and $\ctxtrans{\Phi'}{\Upsilon}{\Gamma}$ and $\eana{\Upsilon}{\Phi'}{e}{\st}{\iota}$ or $\esyn{\Upsilon}{\Phi'}{e}{\st}{\iota}$ then we define $\bprjX{e}{e'}$ by mapping the following transformation over all sub-terms $e''$: \begin{enumerate}
+\item If $e''=\eintro{\sttmidx}{\es}$ such that \begin{enumerate}[(1)]
+  \item $\tcdef{\tc}{\tcsig{\ktyidx}{\chi}}{\tcstruct{\stx{schema}}{\omega}} \in \Phi$
+  \item $\introsig{\klitidx} \in \chi$
+  \item $\sofkz{\emptyset}{\emptyset}{\Phi}{\stmidx}{\klitidx}$
+  \item $\keyw{ana~intro}={\stx{def}} \in \omega$
+  \item $|\es| = n$
+  \item $\keyw{args}(n)=\stx{args}$
+  \item $\stx{def}(\sttyidx)(\stmidx)(\stx{args}) \Downarrow_{\es;\Upsilon;\Phi} \sqitm{\qitm}$
+  \item $\tdeabs{\Phi}{c}{\srep{\st}}{\emptyset}{\ity_\text{abs}}{\memD}$
+  \item $\edeabs{c}{{\es;\Upsilon;\Phi}}{\qitm}{\memD}{\emptyset}{\iota_\text{abs}}{\memD'}{\memG}$
+  \item $\memD' \leadsto \delta : \Delta_\text{abs}$
+  \item $\memG \leadsto \gamma : \Gamma_\text{abs}$
+  \item $\Delta_\text{abs}~\Gamma_\text{abs} \vdash \iota_\text{abs} : \tau_\text{abs}$
+\end{enumerate}
+
+We generate $\eintro{\sqity{\qity}}{\es'}$ where $\bprjX{\es}{\es'}$ maps backprojection recursively over all terms in $\es$ for which a type was recorded in $\memG$.
+All others can be left alone (they have no bearing on the type or translation).
+
+\item If $e''=\etarg{\opname{op}}{\sttmidx}{e_\text{targ}}{\es}$ the procedure is analagous.
+\end{enumerate}
+\end{definition}
+
+Comparing (ana-intro) and (ana-intro-other), and (syn-targ) and (syn-targ-other), it is straightforward to see that term translations are invariant with respect to backprojection, and types respect backprojection.
+
+\begin{lemma}
+If $\vdash \Phi'$ and $\ctxtrans{\Phi'}{\Upsilon}{\Gamma}$ and $\eana{\Upsilon}{\Phi'}{e}{\st}{\iota}$ or $\esyn{\Upsilon}{\Phi'}{e}{\st}{\iota}$ then $\bprjX{e}{e'}$ and $\bprjX{\Upsilon}{\Upsilon'}$ and $\ctxtrans{\Phi}{\Upsilon'}{\Gamma}$ and $\bprjX{\st}{\st'}$ and $\eana{\Upsilon'}{\Phi}{e}{\st'}{\iota}$.
+\end{lemma}
+
+The end result is that we have our conservativity theorem.
+
+\begin{theorem}[Conservativity]\label{thm:conservativity} If $\vdash \Phi$ and $\tc \in \text{dom}(\Phi)$ and a tycon invariant for $\tc$ holds under $\Phi$: \begin{itemize}
+\item For all $\Upsilon, e, \sttyidx$, if $\eana{\Upsilon}{\Phi}{e}{\sty{\tc}{\sttyidx}}{\iota}$ and $\ctxtransX{\Upsilon}{\Gamma}$ and $\vdash_\Phi {\sty{\tc}{\sttyidx}} \leadsto \tau$  then $P(\Gamma, \sttyidx, \iota)$.
+\end{itemize} then for all $\Phi' = \Phi, \tcdef{\tc'}{\tcsig{\kappa'}{\chi'}}{\theta'}$ such that $\vdash \Phi'$, the same tycon invariant holds under $\Phi'$: \begin{itemize}
+\item For all $\Upsilon, e, \sttyidx$, if $\eana{\Upsilon}{\Phi'}{e}{\sty{\tc}{\sttyidx}}{\iota}$ and $\vdash_{\Phi'} \Upsilon \leadsto \Gamma$ and $\vdash_{\Phi'}{\sty{\tc}{\sttyidx}}\leadsto{\tau}$ then $P(\Gamma, \sttyidx, \iota)$.
+\end{itemize}
+(if proposition $P(\Gamma, \st, \iota)$ is \emph{modular})
+\end{theorem}
+\begin{proof-sketch}
+We simply backproject our typing derivation, apply the original tycon invariant, then apply the definition of modular propositions.\end{proof-sketch}
+
+This section has, obviously, been less detailed than the remainder of the supplement, essentially because the lemmas above involve a rather large number of cases where nothing of note is happening, and we simply ran out of time to inductively specify the full details of backprojection and prove all of them.
+Backprojection is a rather simple conceptual mapping, so we believe that the level of detail above is sufficiently convincing, and we plan on writing out the full details (which we anticipate will be another 20 or so pages) well before the notification date.
+
+% \subsection{Decidability}
+% Not going to prove this so probably take it out.
+\section{Discussion}
+\subsection{Record Types}
+In the paper, we presented labeled products rather than the more conventional record types, both to emphasize that different variants have different trade-offs, and because it made our examples simpler.
+
+One strategy for supporting records themselves would be to specify the intro operator such that records that do not have sorted indices are uninhabited, then provide a helper static function that generates only sorted indices and require that all type construction pass through that.
+This notion of a ``type constructor constructor'' could perhaps be integrated into the calculus.
+
+An alternative strategy may have been to use an abstract kind that ensured that such type indices could not have been constructed, but to be compatible with our equality kind restriction, this would require support for abstract equality kinds, analagous to abstract equality types in SML.
+We chose not to formalize these for simplicity, and to demonstrate the general technique above.
+This could also have been used for the labeled product example.
+
+\subsection{Conjectures}
+Other properties that we conjecture are true but have not yet proven include decidability of typing (which requires proving strong normalization of the SL), unicity of typing (i.e.
+that types and translations, if they exist, are unique; this requires proving determinism of normalization), and stability (i.e.
+that extending the tycon context doesn't change the meaning of already well-typed terms; this is essentially trivial by inspection of the rules, and we proved stability of kinding and tycon context validity above because these were needed for other purposes, but ran out of time and space in the paper to even mention this).
+Note that because these have no formal development, we did not mention them as contributions in the paper.
+
+\subsection{Additional Examples}
+We have developed additional examples, and developed detailed proof sketches of tycon invariants with them.
+These will also be available in the final supplement, but due to recent changes in our calculus, and the tedium of writing ``code'' using TeX macros, these also didn't make the cut in the submitted supplement.
+There is nothing conceptually novel in these examples, and we did not claim that we had additional complete examples available upon submission in the paper.
+
+Note that because of our simplification related to inductive kinds here, the regular string example cannot be encoded (kind $\krx$ is an inductive sum).
+
 % We recommend abbrvnat bibliography style.
-
+\include{att-pldi15-supplement-figures}
+
+\clearpage
 \bibliographystyle{abbrv}
@@
 %P. Q. Smith, and X. Y. Jones. ...reference text...
-
-\newpage
-\appendix
-\section{Appendix}
-\begin{mathpar}
-\small
-\inferrule[s-ty-step]{
-    \sstep{\st}{\argEnv}{\st'}
-}{
-    \sstep{\sty{c}{\st}}{\argEnv}{\sty{c}{\st'}}
-}
-
-\inferrule[s-ty-err]{
-    \serr{\st}{\argEnv}
-}{
-    \serr{\sty{c}{\st}}{\argEnv}
-}
-
-\inferrule[s-ty-v]{
-    \sval{\st}{\argEnv}
-}{
-    \sval{\sty{c}{\st}}{\argEnv}
-}
-
-\inferrule[s-otherty-v]{ }{
-    \sval{\sotherty{m}{\tau}}{\argEnv}
-}
-\end{mathpar}
-
-\begin{mathpar}
-\small
-\inferrule[s-tycase-step]{
-    \sstep{\st}{\argEnv}{\st'}
-}{
-    \sstep{\stycase{c}{\st}{\sx}{\st_1}{\st_2}}{\argEnv}{\stycase{c}{\st'}{x}{\st_1}{\st_2}}
-}
-
-\inferrule[s-tycase-err]{
-    \serr{\st}{\argEnv}
-}{
-    \serr{\stycase{c}{\st}{\sx}{\st_1}{\st_2}}{\argEnv}
-}
-\end{mathpar}
-\begin{mathpar}\small
-\inferrule[s-tycase-match]{
-    \sval{\sty{c}{\st}}{\argEnv}
-}{
-    \sstep{\stycase{c}{\sty{c}{\st}}{\sx}{\st_1}{\st_2}}{\argEnv}{[\st/x]\st_1}
-}
-
-\inferrule[s-tycase-fail]{
-    \st \neq \sty{c}{\st'}
-}{
-    \sstep{\stycase{c}{\st}{\sx}{\st_1}{\st_2}}{\argEnv}{\st_2}
-}
-\end{mathpar}
-\begin{mathpar}
-\small
-\inferrule[keq-k]{\kvar \in \kDelta}{\keq{\kDelta}{\kvar}}
-
-\inferrule[keq-ind]{\keq{\kDelta, \kvar}{\kappa}}{
-  \keq{\kDelta}{\kmu{\kvar}{\kappa}}}
-
-\inferrule[keq-unit]{ }{\keq{\kDelta}{\kunit}}
-\end{mathpar}\begin{mathpar}
-\small
-\inferrule[keq-prod]{
-  \keq{\kDelta}{\kappa_1}~~~~
-  \keq{\kDelta}{\kappa_2}
-}{
-  \keq{\kDelta}{\kprod{\kappa_1}{\kappa_2}}
-}
-
-\inferrule[keq-sum]{
-  \keq{\kDelta}{\kappa_1}~~~~
-  \keq{\kDelta}{\kappa_2}
-}{
-  \keq{\kDelta}{\ksum{\kappa_1}{\kappa_2}}
-}
-
-\inferrule[keq-ty]{ }{\keq{\kDelta}{\kty}}
-\end{mathpar}
- %We will discuss strategies for fragments that benefit from more sophisticated equivalences in Section \ref{examples}.\todo{do this}
-
-\begin{mathpar}
-
-\inferrule[k-ity-alpha]{ }{\sofkX{\sqity{\alpha}}{\kity}}
-\end{mathpar}
-
-\begin{mathpar}
-\small
-\inferrule[s-ity-lam-step-1]{
-    \sstep{\sqity{\qity_1}}{\argEnv}{\sqity{\qity_1'}}%\sstep{\sqity{\qity_1}}{\argEnv}{\sstep{\sqity{\qity_1'}}}
-}{
-    \sstep{\sqity{\qity_1\times\qity_2}}{\argEnv}{\sqity{\qity_1' \times \qity_2}}
-}
-
-\inferrule[s-ity-lam-step-2]{
-    \sval{\sqity{\qity_1}}{\argEnv}\\
-    \sstep{\sqity{\qity_2}}{\argEnv}{\sqity{\qity_2'}}%\sstep{\sqity{\qity_1}}{\argEnv}{\sstep{\sqity{\qity_1'}}}
-}{
-    \sstep{\sqity{\qity_1\times\qity_2}}{\argEnv}{\sqity{\qity_1 \times \qity_2'}}
-}
-
-\inferrule[s-ity-lam-err-1]{
-    \serr{\sqity{\qity_1}}{\argEnv}
-}{
-    \serr{\sqity{\qity_1\times\qity_2}}{\argEnv}
-}
-
-\inferrule[s-ity-lam-err-2]{
-    \serr{\sqity{\qity_2}}{\argEnv}
-}{
-    \serr{\sqity{\qity_1\times\qity_2}}{\argEnv}
-}
-
-\inferrule[s-ity-lam-v]{
-    \sval{\sqity{\qity_1}}{\argEnv}\\
-    \sval{\sqity{\qity_2}}{\argEnv}
-}{
-    \sval{\sqity{\qity_1\times\qity_2}}{\argEnv}
-}
-
-\inferrule[s-ity-alpha-v]{ }{
-    \sval{\sqity{\alpha}}{\argEnv}
-}
-\end{mathpar}
-\begin{mathpar}
-\small
-\inferrule[k-tycase-parr]{
-    \sofkX{\st}{\kty}\\
-    \sofk{\kDelta}{\kGamma, \sx :: \kty \times \kty}{\Phi}{\st_1}{\kappa}\\
-    \sofkX{\st_2}{\kappa}
-}{
-    \sofkX{\stycase{\rightharpoonup}{\st}{\sx}{\st_1}{\st_2}}{\kappa}
-}
-\end{mathpar}
-\begin{mathpar}
-\small
-\inferrule[s-ity-unquote-step]{
-    \sstep{\st}{\argEnv}{\st'}
-}{
-    \sstep{\sqity{\qtuq{\st}}}{\argEnv}{\sqity{\qtuq{\st'}}}
-}
-
-\inferrule[s-ity-unquote-err]{
-    \serr{\st}{\argEnv}
-}{
-    \serr{\sqity{\qtuq{\st}}}{\argEnv}
-}
-\end{mathpar}
-\begin{mathpar}
-\inferrule[s-ity-trans-step]{
-    \sstep{\st}{\argEnv}{\st'}
-}{
-    \sstep{\sqity{\srep{\st}}}{\argEnv}{\sqity{\srep{\st'}}}
-}
-
-\inferrule[s-ity-trans-err]{
-    \serr{\st}{\argEnv}
-}{
-    \serr{\sqity{\srep{\st}}}{\argEnv}
-}
-\end{mathpar}
-This judgement is defined by the following straightforward rules:
-\begin{mathpar}\small
-\inferrule[tstore-emp]{ }{\emptyset \leadsto \emptyset : \emptyset}
-
-\inferrule[tstore-ext]{\memD \leadsto \delta : \Delta}{(\memD, \st \leftrightarrow \tau/\alpha) \leadsto (\delta, \tau/\alpha) : (\Delta, \alpha)}
-\end{mathpar}
-
-\[\small
-\begin{array}{lll}
-\textbf{Description} ~~~~& \textbf{Concrete Form}~~~~ & \textbf{Desugared Form}\\
-\text{sequences} & (e_1, \ldots, e_n) ~\text{or}~ \texttt{[}e_1, \ldots, e_n\texttt{]} & \eintro{()}{e_1; \ldots; e_n}\\
-\text{labeled sequences} & \{\mathtt{lbl}_1=e_1, \ldots, \mathtt{lbl}_n=e_n\} & \eintro{\texttt{[}\mathtt{lbl}_1, \ldots, \mathtt{lbl}_n\texttt{]}}{e_1; \ldots; e_n}\\
-\text{label application} & \mathtt{lbl}\langle e_1, \ldots, e_n \rangle & \eintro{\mathtt{lbl}}{e_1, \ldots, e_n}\\
-\text{numerals} & n & \eintro{n}{\cdot}\\
-\text{labeled numerals} & n\texttt{lbl} & \eintro{(n, \texttt{lbl})}{\cdot}\\
-\text{strings} & \texttt{"s"} & \eintro{\texttt{"s"}}{\cdot}
-\end{array}
-\]
-
-\begin{mathpar}\small
-\inferrule[s-itm-var-v]{ }{
-    \sval{\sqitm{x}}{\argEnv}
-}
-
-\inferrule[s-itm-lam-step-1]{
-    \sstep{\sqity{\qity}}{\argEnv}{\sqity{\qity'}}
-}{
-    \sstep{\sqitm{\ilam{\qity}{x}{\qitm}}}{\argEnv}{\sqitm{\ilam{\qity'}{x}{\qitm}}}
-}
-
-\inferrule[s-itm-lam-step-2]{
-    \sval{\sqity{\qity}}{\argEnv}\\
-    \sstep{\sqitm{\qitm}}{\argEnv}{\sqitm{\qitm'}}
-}{
-    \sstep{\sqitm{\ilam{\qity}{x}{\qitm}}}{\argEnv}{\sqitm{\ilam{\qity}{x}{\qitm'}}}
-}
-\end{mathpar}
-\begin{mathpar}\small
-\inferrule[s-itm-lam-err-1]{
-    \serr{\sqity{\qity}}{\argEnv}
-}{
-    \serr{\sqitm{\ilam{\qity}{x}{\qitm}}}{\argEnv}
-}
-
-\inferrule[s-itm-lam-err-2]{
-    \serr{\sqitm{\qitm}}{\argEnv}
-}{
-    \serr{\sqitm{\ilam{\qity}{x}{\qitm}}}{\argEnv}
-}
-
-\inferrule[s-itm-lam-v]{
-    \sval{\sqity{\qity}}{\argEnv}\\
-    \sval{\sqitm{\qitm}}{\argEnv}
-}{
-    \sval{\sqitm{\ilam{\qity}{x}{\qitm}}}{\argEnv}
-}
-\end{mathpar}
-\begin{mathpar}\small
-\inferrule[k-itm-unquote]{
-    \sofkX{\st}{\kitm}
-}{
-    \sofkX{\sqitm{\quq{\st}}}{\kitm}
-}
-\end{mathpar}
-
-\begin{mathpar}\small
-\inferrule[s-itm-unquote-step]{
-    \sstep{\st}{\argEnv}{\st'}
-}{
-    \sstep{\sqitm{\quq{\st}}}{\argEnv}{\sqitm{\quq{\st'}}}
-}
-
-\inferrule[s-itm-unquote-err]{
-    \serr{\st}{\argEnv}
-}{
-    \serr{\sqitm{\quq{\st}}}{\argEnv}
-}
-
-\inferrule[s-itm-unquote-elim]{
-    \sval{\sqitm{\qitm}}{\argEnv}
-}{
-    \sstep{\sqitm{\quq{\sqitm{\qitm}}}}{\argEnv}{\sqitm{\qitm}}
-}
-\end{mathpar}
-\begin{mathpar}
-\inferrule[s-ana-step]{
-    \sstep{\st}{\argEnv}{\st'}
-}{
-    \sstep{\sana{n}{\st}}{\argEnv}{\sana{n}{\st'}}
-}
-
-\inferrule[s-ana-err]{
-    \serr{\st}{\argEnv}
-}{
-    \serr{\sana{n}{\st}}{\argEnv}
-}
-
-\inferrule[s-itm-anatrans-step]{
-    \sstep{\st}{\argEnv}{\st'}
-}{
-    \sstep{\sqitm{\anatrans{n}{\st}}}{\argEnv}{\sqitm{\anatrans{n}{\st'}}}
-}
-
-\inferrule[s-itm-anatrans-err]{
-    \serr{\st}{\argEnv}
-}{
-    \serr{\sqitm{\anatrans{n}{\st}}}{\argEnv}
-}
-\end{mathpar}
-, as specified by the judgement $\memG \leadsto \gamma : \Gamma$ defined by the following rules:
-\begin{mathpar}
-\inferrule[ttrs-emp]{ }{\emptyset \leadsto \emptyset : \emptyset}
-
-\inferrule[ttrs-ext]{
-    \memG \leadsto \gamma : \Gamma
-}{
-    (\memG, n : \st \leadsto \iota/x : \tau) \leadsto (\gamma, \iota/x) : (\Gamma, x : \tau)
-}
-\end{mathpar}
-
-\begin{mathpar}\small
-\inferrule[k-itm-lam]{
-    \sofkX{\sqity{\qity}}{\kity}\\
-    \sofkX{\sqitm{\qitm}}{\kitm}
-}{
-    \sofkX{\sqitm{\ilam{\qity}{x}{\qitm}}}{\kitm}
-}
-\end{mathpar} 
-\begin{mathpar}\small
-\inferrule[k-raise]{\kDelta \vdash \kappa}{\sofkX{\sraise{\kappa}}{\kappa}}
-
-\inferrule[s-raise]{ }{\serr{\sraise{\kappa}}{\argEnv}}
-\end{mathpar}
-
-\begin{mathpar}\small
-\inferrule[s-syn-success]{
-    \keyw{nth}[n](\es) = e\\
-    \esynX{e}{\st}{\iota}
-}{
-    \sstep{\ssyn{n}}{\es;\Upsilon;\Phi}{(\st, \sqitm{\syntrans{n}})}
-}
-
-\inferrule[s-syn-fail]{
-    \keyw{nth}[n](\es) = e\\
-    [\Upsilon \vdash_\Phi e \nRightarrow]
-}{
-    \serr{\ssyn{n}}{\es;\Upsilon;\Phi}
-}
-\end{mathpar}
-
-\begin{mathpar}\small
-\inferrule[k-itm-syntrans]{
-    n' < n
-}{
-    \sofkX{\sqitm{\syntrans{n'}}}{\kitm}
-}
-\end{mathpar}
-, e.g. for lambdas:
-\begin{mathpar}\small
-%\inferrule[abs-var]{ }{\edeabs{\tc}{\argEnv}{x}{\memD}{\memG}{x}{\memD}{\memG}}
-%
-\inferrule[abs-lam]{
-    \tdeabs{\Phi}{\tc}{\qity}{\memD}{\ity}{\memD'}\\
-    \edeabs{\tc}{\es;\Upsilon;\Phi}{\qitm}{\memD'}{\memG}{\itm}{\memD''}{\memG'}
-}{
-    \edeabs{\tc}{\es;\Upsilon;\Phi}{\ilam{\qity}{x}{\qitm}}{\memD}{\memG}{\ilam{\tau}{x}{\itm}}{\memG'}{\memD''}
-}
-\end{mathpar}
-\begin{mathpar}\small
-\inferrule[abs-anatrans-stored]{
-    n : \st \leadsto \iota/x : \tau \in \memG
-}{
-    \edeabs{\tc}{\argEnv}{\anatrans{n}{\st}}{\memG}{\memD}{x}{\memG}{\memD}
-}
-
-\inferrule[abs-syntrans-stored]{
-    n : \st \leadsto \iota/x : \tau \in \memG
-}{
-    \edeabs{\tc}{\argEnv}{\syntrans{n}}{\memG}{\memD}{x}{\memG}{\memD}
-}\end{mathpar}
-\begin{mathpar}\small
-\inferrule[k-itm-anatrans]{
-    n' < n\\
-    \sofkX{\st}{\kty}
-}{
-    \sofkX{\sqitm{\anatrans{n'}{\st}}}{\kitm}
-}
-\end{mathpar}
-\begin{mathpar}\small
-\inferrule[abs-syntrans-new]{
-    n \notin \text{dom}(\memG)\\
-    \keyw{nth}[n](\es) = e\\
-    \esynX{e}{\st}{\iota}\\
-    \tdeabs{\Phi}{\tc}{\srep{\st}}{\memD}{\ity}{\memD'}\\
-    (x~\text{fresh})
-}{
-    \edeabs{\tc}{\es;\Upsilon;\Phi}{\syntrans{n}}{\memG}{\memD}{x}{\memG, n : \st \leadsto \iota/x : \tau}{\memD'}
-}
-\end{mathpar}
-
-\begin{mathpar}\small
-\inferrule[etctx-emp]{ }{\vdash_\Phi \emptyset \leadsto \emptyset}
-
-\inferrule[etctx-ext]{\vdash_\Phi \Upsilon \leadsto \Gamma\\\istype{\st}{\Phi}\\\vdash_\Phi \st \leadsto \tau}{\vdash_\Phi \Upsilon, x \Rightarrow \st \leadsto \Gamma, x : \tau}
-\end{mathpar}
-
-\[\small
-\begin{array}{lll}
-\textbf{Description}& \textbf{Concrete Form}~~~~ & \textbf{Desugared Form}\\
-\text{index projection} & e_\text{targ}\texttt{\#}n & \etarg{\opname{idx}}{n}{e_\text{targ}}{\cdot}\\
-\text{label projection} & e_\text{targ}\texttt{\#}\conclbl{lbl} & \etarg{\opname{prj}}{\conclbl{lbl}}{e_\text{targ}}{\cdot}\\
-\text{explicit invocation}~~~~ & e_\text{targ}{\cdot}\opname{op}[\sttmidx](\es) & \etarg{\opname{op}}{\sttmidx}{e_\text{targ}}{\es}\\
-& e_\text{targ}{\cdot}\opname{op}(\es) & \etarg{\opname{op}}{()}{e_\text{targ}}{\es}\\
-& e_\text{targ}{\cdot}\opname{op}(\conclbl{lbl}_1=e_1, \ldots, \conclbl{lbl}_n=e_n) & \keyw{targ}[\opname{op}; \texttt{[}\conclbl{lbl}_1, \ldots, \conclbl{lbl}_n\texttt{]}](e_\text{targ}; \\
-& & \quad e_1; \ldots; e_n)\\
-\text{labeled case analysis} & e_\text{targ}{\cdot}\opname{case}\,\{ & \keyw{targ}[{\opname{case}}; {\texttt{[}\st_1, \ldots, \st_n\texttt{]}}]({e_\text{targ}}; \\
- & ~|~~\st_1\langle x_1, \ldots, x_k \rangle \Rightarrow e_1 & \quad \eanalam{x_1}{\ldots \eanalam{x_k}{e_1}};\\
- & ~|~~\ldots & \quad \ldots;\\ 
- & ~|~~\st_n\langle x_1, \ldots, x_k \rangle \Rightarrow e_n\} & \quad \eanalam{x_1}{\ldots \eanalam{x_k}{e_n}})
-\end{array}
-\]
-
- For example, 
-\begin{mathpar}
-\small
-\inferrule[abs-prod]{
-    \tdeabs{\Phi}{\tc}{\qity_1}{\memD}{\ity_1}{\memD'}\\
-    \tdeabs{\Phi}{\tc}{\qity_2}{\memD'}{\ity_2}{\memD''}
-}{
-    \tdeabs{\Phi}{\tc}{\qity_1 \times \qity_2}{\memD}{\ity_1 \times \ity_2}{\memD''}
-}
-\end{mathpar}
-
-The argument interfaces that populate the list provided to opcon definitions is derived from the argument list by the judgement $\keyw{args}(\es)=_n \stx{args}$, defined as follows:
-\begin{mathpar}\small
-\inferrule[args-z]{ }{
-    \keyw{args}(\cdot)=_0 \skap{\karg}{\svar{nil}}
-}
-
-\inferrule[args-s]{
-    \keyw{args}(\es)=_{n} \st
-}{
-    \keyw{args}(\es;e)=_{n+1} \skap{\karg}{\svar{rcons}}~\st~(\slam{\kty}{\svar{ty}}{\sana{n}{\svar{ty}}},\slam{\kunit}{\_}{\ssyn{n}})
-}
-\end{mathpar}
-We assume that the definitions of the standard helper functions $\small\svar{nil} :: \kforall{\kalpha}{\klist{\kalpha}}$ and $\small\svar{rcons} :: \kforall{\kalpha}{\klist{\kalpha} \rightarrow \kalpha \rightarrow \klist{\kalpha}}$, which adds an item to the end of a list, have been substituted into these rules.
-The result is that the $n$th element of the argument interface list simply wraps the static terms $\sana{n}{\st}$ and $\ssyn{n}$.\end{document}
+\end{document}

### 2014-11-16 att-pldi15/att-pldi15.tex
final PLDI15 submission

@@
 %\documentclass[12pt]{article}
-\documentclass[10pt,preprint]{sigplanconf}
+\documentclass[pldi]{sigplanconf-pldi15}
 % The following \documentclass options may be useful:
@@
 
+\usepackage[colorlinks=true,allcolors=blue,breaklinks,draft=false]{hyperref}   % hyperlinks, including DOIs and URLs in bibliography
+% known bug: http://tex.stackexchange.com/questions/1522/pdfendlink-ended-up-in-different-nesting-level-than-pdfstartlink
+
 \begin{document}
@@
 \begin{abstract}
-Researchers often describe type systems as fragments or simple calculi, leaving to language designers the task of composing these to form complete programming  languages.
-%comparable to those available for library-based embeddings packaged using a modern module system.
-This is not a systematic process: metatheoretic results must be established anew for each composition, guided only notionally by metatheorems derived for simpler systems.
-As the language design space grows, mechanisms that provide stronger modular reasoning principles than this are needed.%, so that important metatheoretic results need only be established locally and, ultimately, so that language extensions can be treated like libraries.
-%,  Sometimes this fails because certain language features do not ``play well'' together, but precisely characterizing which is elusive given these practices.
-%Recent work has started to address this problem, e.g.
-by showing how to modularly reason about desugarings atop a fixed type system.
-Our focus here is on safely composing type system fragments.
-We organize fragments around type constructors, as is usual practice when describing type systems, and sidestep the difficulties of composing abstract syntax (i.e.
-the expression problem)  by delegating control over a small, uniform abstract syntax in a type-directed manner to logic associated with a type constructor.
+Researchers often describe type systems as fragments, leaving to language designers the task of composing these to form complete programming  languages. 
+This has not been a systematic process: metatheoretic results must be established anew for each composition of fragments, guided only informally by metatheorems derived for simpler systems.
+As the language design space grows, mechanisms that provide stronger modular reasoning principles than this are needed.
 
@@
 row projection) using a static language where types and translations are values.
-The semantics come with strong metatheoretic guarantees, notably \emph{type safety} and \emph{conservativity}: that \emph{tycon-specific  invariants} established in any ``closed world'' are conserved in the ``open world''.
+The semantics come with modular metatheoretic guarantees, notably \emph{type safety} and \emph{conservativity}: that \emph{tycon-specific  invariants} established in any ``closed world'' are conserved in the ``open world''.
 Remarkably, extension providers do not need to  provide the semantics with mechanized specifications or proofs.
-Instead, these guarantees arise from a form of translation validation that, taking inspiration from the ML module system, uses type abstraction to check that the opcons associated with a tycon respect the \emph{translation independence} of all others.% We can thus call them \emph{modular tycons}.
+Instead, these guarantees arise from a form of translation validation that, taking inspiration from the ML module system, uses type abstraction to check that the opcons associated with a tycon respect the \emph{translation independence} of all others.
 \end{abstract}
@@
 \section{Introduction}\label{intro}
-Typed programming languages are generally described as being composed from \emph{fragments}, each contributing to the language's concrete syntax, abstract syntax, static semantics and dynamic semantics.
-In his textbook, Harper organizes such fragments around type constructors, describing each in a different chapter \cite{pfpl}.
-Languages are then identified by a set of these type constructors, e.g.
-$\mathcal{L}\{{\rightharpoonup}\,{\forall}\,{\mu}\,{\keyw{1}}\,{\times}\,{+}\}$ is the language of partial function types, polymorphic types, recursive types, nullary and binary product types and binary sum types (its syntax is shown in Figure \ref{syntax-IL}, discussed below).
-Another common practice is to describe a fragment using a simple calculus having a ``catch-all'' constant and base type to stand notionally for all other terms and types that may also be included in some future complete language (e.g.
+Typed programming languages are usually described as consisting of \emph{fragments}, each contributing separately to the concrete and abstract syntax and static and dynamic semantics of the language \cite{pfpl}.
+Harper organizes such fragments around \emph{type constructors}, e.g.
+$\rightarrow$ for total function types, describing each in a different chapter of his book.
+Languages are then identified by a set of type constructors, e.g.
+$\mathcal{L}\{{\rightharpoonup}\,{\forall}\,{\mu}\,{\keyw{1}}\,{\times}\,{+}\}$ is the language of partial (i.e.
+possibly non-terminating) function types, polymorphic types, recursive types, nullary and binary product types and binary sum types (its syntax is shown in Figure \ref{syntax-IL}, discussed below).
+Another common practice is to describe a fragment using a simple calculus having a ``catch-all'' constant and base type to stand notionally for all ``other'' terms and types.% (e.g.
 \cite{sanitation-psp14}).
@@
 This is not an everyday problem for programmers only because fragments like those mentioned above are ``general purpose'': they make it possible to \emph{isomorphically embed}  many other fragments as ``libraries''.
-For example, list types need not be built in because they are isomorphic to the type $\iforall{\alpha}{\imu{t}{\iunit + (\alpha \times t))}}$ (datatypes in ML combine these into a single declaration construct).
+For example, a list type constructor need not be built in because lists are isomorphic to the type $\iforall{\alpha}{\imu{t}{\iunit + (\alpha \times t))}}$ (datatypes in ML combine these into a single declaration construct).
 %Languages providing datatypes al\'a ML  are perhaps most directly oriented around  embeddings into this core.
@@
 
-Universality properties (e.g.
-``Turing completeness'') are often enough to guarantee that an embedding that preserves a desirable fragment's dynamic semantics can be constructed, but an isomorphic embedding must also preserve the static semantics and, if defined, performance bounds specified by a cost semantics.
-This is not always possible.
-For example, in $\mathcal{L}\{{\rightharpoonup}\,{\forall}\,{\mu}\,{\keyw{1}}\,{\times}\,{+}\}$, it is impossible to introduce record types as a library because this requires introducing row projection operators,  written $\opname{\#}\conclbl{lbl}$ in ML, one for each of the infinite set of row labels $\conclbl{lbl}$ ($\opname{\#}$  is thus an \emph{operator constructor}).
+%But while universality properties (e.g. ``Turing completeness'') are often enough to guarantee that an embedding that preserves a desirable fragment's dynamic semantics can be constructed, 
+Establishing an isomorphic embedding of a fragment is not always possible, however, because it requires preserving both the static and dynamic semantics and, if defined, performance bounds specified by a cost semantics, of all operators.
+This is possible if they have ``function-like'' semantics, as the list operators do.
+But in $\mathcal{L}\{{\rightharpoonup}\,{\forall}\,{\mu}\,{\keyw{1}}\,{\times}\,{+}\}$, it is impossible to introduce record types as a library because this requires introducing row projection operators,  written $\opname{\#}\conclbl{lbl}$ in ML, one for each of the infinite set of row labels $\conclbl{lbl}$ ($\opname{\#}$  is thus an \emph{operator constructor}).
 Each time such a situation occurs,  a new \emph{dialect} is needed.
-The ML lineage, like others, faces a proliferation of dialects:%, showing that this is not an isolated example:
+The ML lineage, like others, faces a proliferation of dialects that go beyond ``core ML'':%, showing that this is not an isolated example:
 %Reynolds, in a remark that recalls the ``Turing tarpit'' of Perlis \cite{Perl82a}, summarizes the issue \cite{Reynolds94anintroduction}: 
@@
 records (identified up to reordering), and 
-records with width and depth subtyping \cite{Cardelli:1984:SMI:1096.1098}, functional update operators\footnote{The Haskell wiki notes that ''No, extensible records are not implemented in GHC.
+records with width and depth subtyping \cite{Cardelli:1984:SMI:1096.1098} and functional update operators\footnote{The Haskell wiki notes that ''No, extensible records are not implemented in GHC.
 The problem is that the record design space is large, and seems to lack local optima.
-[...] As a result, nothing much happens.'' \cite{GHCFAQ}} \cite{ocaml-manual},
- mutable fields \cite{ocaml-manual}, 
+[...] As a result, nothing much happens.'' \cite{GHCFAQ}} \cite{ocaml-manual}%,
+ %mutable fields \cite{ocaml-manual}, 
 %field delegation \cite{atlang-gpce14} \todo{gpce submission} 
-and 
- ``methods'' (i.e. pure objects) \cite{TSLs}. 
+%and 
+% ``methods'' (i.e. pure objects) \cite{TSLs}
+. 
  Sum-like types are also exposed variously: standard datatypes, 
@@
 polymorphic variants \cite{ocaml-manual} and 
-exception types \cite{harper1997programming}. Combinations occur as class-based object systems \cite{ocaml-manual}.
+exception types \cite{Tofte:89:TheDefinitionOfStandardML}. Combinations occur as class-based object systems \cite{ocaml-manual}.
 
 \item
-\textbf{Specialized Fragments:} Fragments providing more specialized operators are also frequently introduced in dialects, e.g.
-for distributed programming \cite{Murphy:2007:TDP:1793574.1793585}, reactive programming \cite{mandel2005reactiveml}, databases \cite{Ohori:2011:MSM:2034773.2034815},  units of measure \cite{conf/cefp/Kennedy09} and regular string sanitation \cite{sanitation-psp14}, amongst many others.
-%\verb|sprintf| in the OCaml dialect statically distinguishes format strings from strings.% All of these are implemented as dialects of existing languages, presumably because a strong encoding was not feasible.
+\textbf{Specialized Fragments:} Fragments providing more specialized operators are also commonly introduced in dialects, e.g.
+for distributed programming \cite{Murphy:2007:TDP:1793574.1793585}, concurrency \cite{AliceLookingGlass}, reactive programming \cite{mandel2005reactiveml}, databases \cite{Ohori:2011:MSM:2034773.2034815},  units of measure \cite{conf/cefp/Kennedy09} and regular string sanitation \cite{sanitation-psp14}.%\verb|sprintf| in the OCaml dialect statically distinguishes format strings from strings.% All of these are implemented as dialects of existing languages, presumably because a strong encoding was not feasible.
 \vspace{-3px}
 \item
-\textbf{Foreign Fragments:} A safe and natural foreign function interface (FFI) can be valuable (particularly given this proliferation of dialects).
+\textbf{Foreign Fragments:} A safe and natural foreign function interface (FFI) can be valuable.
 This requires enforcing the type system of the foreign language in the calling language.
@@
 
-These problems do not arise for fragment expressed as an isomorphic embedding (i.e.
+These problems with composition do not arise when a fragment is expressed as an isomorphic embedding (i.e.
 as a library) because modern \emph{module systems} can enforce abstraction barriers that ensure that the isomorphism needs only to be established in the ``closed world'' of the module.
-For example, a module defining sets in ML can hold the representation of sets abstract, ensuring that any invariants necessary to maintain the isomorphism need only be maintained by the functions in the module (e.g.
+For example, a module for sets in ML can hold its representation type abstract, ensuring that any invariants necessary to maintain the isomorphism need only be maintained by the functions in the module (e.g.
 uniqueness, if using a list representation).
-These then continue to hold no matter which other modules are in use by a client \cite{harper1997programming}.
-Other languages expose similar forms of \emph{abstract data types} that also localize reasoning \cite{liskov1974programming}.
+These then continue to hold no matter which other modules are in use by a client \cite{harper1997programming}.% Other languages expose \emph{abstract data types} in other ways, also to localize reasoning \cite{liskov1974programming}.
 %Mechanisms that can help decrease the complexity of an embedding without violating abstraction barriers are thus valuable, and we will lead into our work in Sec.
@@
 The alarming proliferation of dialects above suggests that mechanisms  that make it possible to define and  reason in a similarly modular, localized manner about  direct extensions to the semantics of a language are needed.
-For example, if a language is extended with \emph{regular string types} as described in \cite{sanitation-psp14}, all terms having a regular string type like $\sty{\tcvar{rstr}}{\concrx{.+}}$ should continue to behave as non-empty strings no matter which other extensions are in use.
+For example, once a language is extended with \emph{regular string types} as described in \cite{sanitation-psp14}, all terms having a regular string type like $\sty{\tcvar{rstr}}{\concrx{.+}}$ should continue to behave as non-empty strings no matter which other extensions are in use.
 %Such a mechanism could ultimately be integrated directly into the language, blurring the distinction between fragments and libraries and decreasing the need for new dialects.% Importing fragments that introduce new semantics   would be as easy and reasonable as importing a new module is in ML today.
@@
 
-\paragraph{Contributions} In this paper, we take foundational steps toward this goal by  constructing a simple but surprisingly powerful core calculus, @$\lambda$ (the ``actively typed'' lambda calculus).
+\paragraph{Contributions} In this paper, we take foundational steps toward this goal of modular language metatheory by  constructing a simple but surprisingly powerful  calculus, @$\lambda$ (the ``actively typed'' lambda calculus).
 %Despite its minimality, it can host a variety of practical semantic extensions like those described above, while maintaining strong metatheoretic guarantees and, crucially, providing modular reasoning properties.
@@
 \ref{overview}.
-Its semantics are structured like those of many modern languages, consisting of an \emph{external language} (EL) governed by a {typed translation semantics} targeting a much simpler \emph{internal language} (IL).
-Rather than building in a monolithic set of external type constructors, however, the  semantics are indexed by a \emph{tycon context}.
+Its semantics consist of an extensible \emph{external language} (EL) governed by a {typed translation semantics} targeting a simple and fixed \emph{internal language} (IL). 
+In particular, rather than building in a monolithic set of external type and operator constructors, the  semantics are indexed by a \emph{tycon context}.
 Each tycon, e.g.
-$\tcvar{lprod}$ defining labeled products or $\tcvar{rstr}$ defining regular strings, determines the semantics of its associated opcons, e.g.
-$\opname{\#}$ for row projection, or $\opname{conc}$ for concatenation, via \emph{static functions}, i.e.
+$\tcvar{lprod}$ defining labeled products, directly specifies the semantics of its associated opcons, e.g.
+$\opname{\#}$ for row projection, via \emph{static functions}, i.e.
 functions written in a \emph{static language} (SL), where types and translations are values.
-In other words, the semantics are controlled by a form of static code generation.
-
-We can call a tycon associated with opcons like this a \emph{modular tycon} because all translation invariants maintained by the opcons associated with a tycon in any ``closed world'', e.g.
-the regular string invariant just mentioned under the tycon context defining only $\tcvar{rstr}$, are necessarily maintained in any further extended tycon context, i.e.
-in the ``open world'', due to a simple static check that maintains \emph{translation independence} between tycons using type abstraction in the IL, the same fundamental principle underlying representation independence in ML-style module systems.
-As in ML, mechanized specifications and proofs are not needed: the modularity guarantee is for all invariants.%, which provides a form of type-level computation.
+%In other words, the semantics are controlled by static \emph{code and type generation}.
+
+We call a tycon associated with opcons in this manner a \emph{modular tycon} because all \emph{translation invariants} maintained by the opcons associated with that tycon, e.g.
+the regular string invariant just mentioned, are necessarily maintained in any further extended tycon context, i.e.
+in the ``open world'', due to a remarkably simple translation validation procedure that maintains \emph{translation independence} between tycons using type abstraction in the IL.
+This is, encouragingly, the same fundamental principle underlying representation independence in ML-style module systems.
+As in ML, mechanized specifications and proofs are not needed.%, which provides a form of type-level computation.
 %The authors demonstrate the expressive power of this technique primarily with an implementation and a number of examples of fragments as libraries, and their core calculus .% (unlike systems that treat the type system as a ``bag of rules'', where non-determinism can arise).
@@
 %Internal terms, $\iota$, together with internal types, $\tau$, form the \emph{typed internal language}. 
-Our approach requires a typed internal language supporting type abstraction (i.e.
+At the heart of our semantics is a typed internal language supporting type abstraction (i.e.
 universal quantification over types) \cite{Reynolds94anintroduction}.
-We use {$\mathcal{L}\{{\rightharpoonup}\,{\forall}\,{\mu}\,{\iunit}\,{\times}\,{+}\}$}, the syntax for which is shown in Figure \ref{syntax-IL}, as representative of a typical intermediate language for a typed language.
+We use {$\mathcal{L}\{{\rightharpoonup}\,{\forall}\,{\mu}\,{\iunit}\,{\times}\,{+}\}$}, Figure \ref{syntax-IL}, as representative of a typical intermediate language for a typed language.
 %Note that the IL has a fixed semantics; changes to the IL induce different dialects of @$\lambda$.
-
-We assume the statics of the IL are specified in the standard way by judgements for  type formation {$\iDelta \vdash \tau$}, typing context formation { $\iDelta \vdash \iGamma$} and type assignment {$\iDelta~\iGamma \vdash \iota : \tau\moutput$}.
+We assume an internal statics specified by judgements for type assignment {$\iDelta~\iGamma \vdash \iota : \tau\moutput$}, type formation {$\iDelta \vdash \tau$} and typing context formation { $\iDelta \vdash \iGamma$}, and an
 %The typing and type formation contexts obey standard structural properties (i.e. weakening, exchange and contraction). 
-The internal dynamics are specified as a structural operational semantics with a stepping judgement {\small $\iota \mapsto \iota\moutput$} and a value judgement {$\iota~\mathtt{val}$}.
-Both the static and dynamic semantics of the IL can be found in any standard textbook covering typed lambda calculi (we directly follow \cite{pfpl}), so we assume familiarity and omit the details.
-
-\paragraph{External Language}
-Programmers interface with @$\lambda$ by writing \emph{external terms}, $e$.
-The abstract syntax of external terms is shown in Figure \ref{syntax-EL} and we will introduce various concrete desugarings in Sec.
-\ref{external-terms}.
+internal dynamics specified as a structural operational semantics with a stepping judgement {\small $\iota \mapsto \iota\moutput$} and a value judgement {$\iota~\mathtt{val}$}.\footnote{Our specifications are intended to be algorithmic: we indicate ``outputs'' when introducing judgement forms by \emph{mode annotations}, $\moutput$.} Both the static and dynamic semantics of the IL can be found in any standard textbook covering typed lambda calculi (e.g.
+\cite{pfpl} or \cite{tapl}), so we assume familiarity.
+
+\paragraph{External Language} Programs ``execute'' as internal terms, but programmers interface with @$\lambda$ by writing \emph{external terms}, $e$.
+The abstract syntax of external terms is shown in Figure \ref{syntax-EL} and we introduce various concrete desugarings as we go on.
 %We will describe useful syntactic desugarings atop this syntax as we go on and review recent techniques that permit modularly introducing  desugarings like these in Sec.
@@
 The semantics are specified as a \emph{bidirectionally typed translation semantics}, i.e.
-the key judgements have the form:\[\esynX{e}{\st\moutput}{\iota\moutput} ~~~~~\text{and}~~~~~ \eanaX{e}{\st}{\iota\moutput}\]
+the key judgements have the following form, pronounced ``Under typing context $\Upsilon$ and tycon context $\Phi$, $e$ (synthesizes / analyzes against) type $\sigma$ and has  translation $\iota$'':\[\esynX{e}{\st\moutput}{\iota\moutput} ~~~~~\text{and}~~~~~ \eanaX{e}{\st}{\iota\moutput}\]
 \noindent
-These are pronounced ``$e$ (synthesizes / analyzes against) type $\sigma$ and has  translation $\iota$ under typing context $\Upsilon$ and tycon context $\Phi$''.
-Our specifications are intended to be algorithmic: we indicate ``outputs'' when introducing judgement forms by \emph{mode annotations}, $\moutput$.
-Note that the type is an ``output'' only for the synthetic judgement.%, i.e.
-it can only be derived when the term itself has enough information in it to determine its type.
-
-We choose bidirectional typechecking, also sometimes called \emph{local type inference} \cite{Pierce:2000:LTI:345099.345100}, for two main reasons.
-The first is once again to justify the practicality of our approach: local type inference is increasingly being used in modern languages (e.g.
-Scala \cite{OdeZenZen01}) because it eliminates the need for type annotations in many situations while being simpler and decidable in more situations than whole-function type inference and providing what are widely perceived to be higher quality error messages \cite{journals/jfp/JonesVWS07}.
-Secondly, it will  give us a clean way to reuse the generalized intro form $\eintro{\st}{\es}$ and its desugarings at many types \cite{TSLs}.
-For example, regular string types will use standard string literal syntax and variants of record types can share   forms like $\{\conclbl{lbl1}=e_1, \conclbl{lbl2}=e_2\}$.
-
-Elaboration semantics distinguishing the EL from the IL have previously been found suitable for full-scale language specifications, e.g.
-the Harper-Stone semantics for Standard ML \cite{Harper00atype-theoretic}.
-The IL is purposely kept small, e.g.
-defining only binary products, to simplify metatheoretic reasoning and compilation.
-The EL then specifies various useful higher-level constructs, e.g.
+%Note that the type is an ``output'' only for the synthetic judgement.%, i.e. it can only be derived when the term itself has enough information in it to determine its type.
+We distinguish situations where the type is an ``output'' from those where it must be provided as an ``'input'' using such a bidirectional approach, also known as \emph{local type inference} \cite{Pierce:2000:LTI:345099.345100}, for two reasons.
+The first is to justify the practicality of our approach: local type inference is increasingly being used in modern languages (e.g.
+Scala \cite{OdeZenZen01}) because it eliminates the need for type annotations in many places and provides high quality error messages.
+%because it eliminates the need for type annotations in many places while being simpler than whole-function type inference and providing what are widely perceived to be higher quality error messages \cite{journals/jfp/JonesVWS07}.
+Secondly, it will  give us a clean way to reuse the abstract introductory form, $\eintro{\st}{\es}$, and its associated desugarings, at many types.% For example, regular string types will use standard string literal syntax, e.g.
+$\concstr{EXMPL}$, and variants of record types will share   forms like $\{\conclbl{lbl1}=e_1, \conclbl{lbl2}=e_2\}$.
+Their meaning is determined by the type they are being analyzed against.
+
+\begin{figure}[t]
+\small
+\hspace{-5px}\input{syntax-table-EL}
+\caption{Abstract syntax of the external language (EL).}
+\label{syntax-EL}\vspace{-10px}
+\end{figure}
+
+For example, consider the following types:
+\[\begin{array}{lcl}
+\stx{title} & := & \sty{\tcvar{rstr}}{\concrx{.+}}\\
+\stx{conf} & := & \sty{\tcvar{rstr}}{\concrx{[A-Z]+ \digit\digit\digit\digit}}\\
+\stx{paper} & := & \sty{\tcvar{lprod}}{\{\conclbl{title} : \stx{title}, \conclbl{conf} : \stx{conf}\}}\end{array}\]
+
+The \emph{regular string types} $\stx{title}$ and $\stx{conf}$ classify values that behave as strings known to be in a specified regular language \cite{sanitation-psp14}, i.e.
+$\stx{title}$ classifies {non-empty strings} and $\stx{conf}$ classifies strings having the format of a typical conference name.
+The \emph{labeled product type} $\stx{paper}$ then describes a conference paper by defining two \emph{rows}, each having one of the regular string types just described.
+Regular string types are defined by  tycon context $\Phi_\text{rstr}$ and labeled products by $\Phi_\text{lprod}$, both introduced in Sec.
+\ref{types}.%Neither are built into the language.
+
+We next define a function, $e_\text{ex}$, that takes a paper title and produces a paper in a conference named $\small\concstr{EXMPL 2015}$: %We define the following concrete terms and types: 
+\[\hspace{-5px}\begin{array}{lcl}
+e_\text{ex} & := & \elam{\stx{title}}{title}{(e_\text{paper} : \stx{paper})}\\
+e_\text{paper} & := & \{\conclbl{title}{=}title, \conclbl{conf}{=}\texttt{"EXMPL 2015"}\}
+\end{array}\]
+
+We will detail the syntax and semantics in Sec.
+\ref{external-terms}.
+To briefly summarize: because of the type ascription $\stx{paper}$ in $e_\text{ex}$, semantic control over $e_\text{paper}$ will be delegated to the \emph{intro opcon definition} of $\tcvar{lprod}$.
+It will decide to analyze the the row value of $\conclbl{title}$ against $\stx{title}$ and the row value of $\conclbl{conf}$, a string literal, against $\stx{conf}$, causing control to pass similarly to $\tcvar{rstr}$.
+Satisfied that the term is well-typed, these will generate translations.
+Thus, we will be able to derive:
+\[\hspace{-5px}\begin{array}{l}\esyn{\emptyset}{\Phi_\text{rstr}\Phi_\text{lprod}}{e_\text{ex}}{(\stx{title}\rightharpoonup\stx{paper})}{\iota_\text{ex}}\end{array}\]
+where $\small\iota_\text{ex} := \ilam{\keyw{str}}{title}{(title, (\texttt{"EXMPL 2015"}_\text{IL}, ()))}$.
+Note that labels need not appear, and $\small\texttt{"EXMPL 2015"}_\text{IL}$ is an internal string (of internal type $\keyw{str}$, defined suitably).
+The trailing unit value arises only because it simplifies our exposition.
+The type annotation on the internal function could be determined because types also have translations, specified by the \emph{type translation judgement} $\vdash_\Phi \st~\mathtt{type} \leadsto \tau\moutput$, read ``$\st$ is a type under $\Phi$ with translation $\tau$''.
+For example, $\stx{title}$  and $\stx{conf}$ have type translations  $\keyw{str}$ and  $\stx{paper}$ in turn has $\keyw{str}\times(\keyw{str}\times\kunit)$.
+Selectively holding type translations abstract will be the key to modular metatheoretic reasoning.
+For example, $\tcvar{lprod}$ cannot claim that a row has a regular string type but produce a translation inconsistent with this claim: the translation $(\concstr{}_\text{IL}, (\concstr{EXMPL}_\text{IL}, ())$ is invalid for a term of type $\stx{paper}$, despite being of internal type $\keyw{str}\times(\keyw{str}\times\kunit)$.
+In fact, it will be checked against $\alpha_1 \times (\alpha_2 \times \kunit)$ (Sec.
+\ref{sec:translation-validation}).
+
+ %Type translations are discussed in Sec.
+\ref{sec:type-translations}.
+%The key to our main result is that a type translation only directly affects the operators associated with it's tycon; all others must be \emph{translation independent}.
+For example, if regular string types had different translations, labeled products would be unaffected.
+%Note however that if either type chose a different type translation, our approach guarantees that the remaining type translations would still be %External typing contexts, $\Upsilon$, map variables to types, so we also need the judgement $\vdash_\Phi \Upsilon~\mathtt{ctx} \leadsto \Gamma$.
+%For example, regular string types will translate to internal string types, abbreviated $\keyw{str}$, and labeled product types will translate to nested binary product types, though we will emphasize that there are other valid choices, and that the choice of type translation should have only local impact.
+
+ %As in typed compilers, the key property implying type safety is that if  $e$ has type $\st$ and translation $\iota$, and $\st$ has translation $\tau$, then $\iota$ must have type $\tau$ (Sec.
+\ref{metatheory}).
+
+ %The IL is purposely kept small, e.g.
+defining only binary products, to simplify metatheoretic reasoning and compiler optimization.
+The EL then specifies useful higher-level constructs, e.g.
 record types, by translation to the IL.
-In @$\lambda$, the EL builds in only function types.
-All other external constructs will arise from the \emph{tycon context}, $\Phi$.
-
-This specification style is perhaps even more directly comparable to a specification for the initial stage of a type-directed compiler, e.g.
-the TIL compiler for Standard ML \cite{tarditi+:til-OLD}, here lifted ``one level up'' into the semantics of the language itself.
-The reason is that in the Harper-Stone elaboration semantics, external and internal terms were governed by a common type system, but in @$\lambda$ each type, $\st$, maps onto an internal type, $\tau$, called the \emph{type translation} of $\st$.
-This mapping is specified by the  type translation judgement, $\vdash_\Phi \st \leadsto \tau$, which will be described in Sec.
-\ref{sec:type-translations}.
-For example, regular string types will translate to internal string types, abbreviated $\keyw{str}$, and labeled product types will translate to nested binary product types, though we will emphasize that there are other valid choices, and that the choice of type translation should have only local impact.
+%In @$\lambda$, the EL builds in only function types. All others arise from the \emph{tycon context}, $\Phi$.
+
+%The reason is that in the Harper-Stone elaboration semantics, external and internal terms were governed by a common type system, but in @$\lambda$
+
  %This will be maintained using a form of translation validation, which also has historically been studied in the literature on compilers \cite{Pnueli-Siegel-Singerman98}.
 
-External typing contexts, $\Upsilon$, map variables to types, so we also need the judgement $\vdash_\Phi \Upsilon \leadsto \Gamma$.
+%External typing contexts, $\Upsilon$, map variables to types, so we also need the judgement $\vdash_\Phi \Upsilon \leadsto \Gamma$.
 
@@
 
-\begin{figure}[t]
-\small
-\hspace{-5px}\input{syntax-table-EL}
-\caption{Syntax of the external language (EL).}
-\label{syntax-EL}\vspace{-5px}
-\end{figure}
-
 % We also here define a syntax for simultaneous substitutions, of terms for variables, $\gamma$, and of types for polymorphic type variables, $\delta$, and assume standard judgements ensuring that these are valid with respect to a valid context, $\Delta \vdash \gamma : \Gamma$ and $\vdash \delta : \Delta$.
@@
 \hspace{-5px}\input{syntax-table-SL}
-\caption{Syntax of the static language (SL). Metavariable $\sx$ ranges over static term variables, $\kalpha$ and $\kvar$ over kind variables and $n$ over natural numbers.}\vspace{-8px}
+\caption{Syntax of the static language (SL). Metavariables $\sx$ ranges over static term variables, $\kalpha$ and $\kvar$ over kind variables and $n$ over natural numbers.}\vspace{-10px}
 \label{syntax-SL}
 \end{figure}
-The main novelty of @$\lambda$ is the \emph{static language} (SL), which itself forms a typed lambda calculus where  
+As suggested above, the main novelty of @$\lambda$ is that the types and term and type translations do not arise from a fixed specification.
+Rather, they are \emph{statically computed} by tycon definitions using a \emph{static language} (SL).
+The SL is itself a typed lambda calculus where
 %External terms are classified by (external) \emph{types}. 
 \emph{kinds}, $\kappa$, serve as the ``types'' of \emph{static terms}, $\sigma$.
-The syntax of the SL is given in Figure \ref{syntax-SL}.
-The portion of the SL covered by the first row of kinds and static terms, some of which are elided for concision, forms an entirely standard total functional programming language consisting of total functions, polymorphic kinds, inductive kinds, and products and sums  \cite{pfpl}.
-The reader can consider these as forming a total subset of ML \cite{harper1997programming} or a simply-typed subset of Coq \cite{Coq:manual} and we will assume standard conveniences in examples (e.g.
-let bindings and inference of type parameters) for concision.
-Only three new  kinds are needed for the SL to serve its role as the language used to   control typing and translation: $\kty$, classifying types (Sec.
-\ref{types}), $\kity$, classifying \emph{quoted translational internal types} (used to compute type translations in Sec.
-\ref{sec:type-translations}) and $\kitm$, classifying \emph{quoted translational internal terms} (used to compute term translations in Sec.
-\ref{sec:introop}).
-The forms $\sana{n}{\st}$ and $\ssyn{n}$ will serve to link the dynamics of the SL with external type synthesis and analysis (Sec.
-\ref{external-terms}).
-
-The kinding judgement takes the form $\sofkX{\st}{\kappa\moutput}$, where $\kDelta$ and $\kGamma$ are analagous to $\iDelta$ and $\iGamma$ and analagous kind and kinding context formation judgements $\kDelta \vdash \kappa$ and $\kDelta \vdash \kGamma$ are  defined.
-All such contexts in @$\lambda$ are identified up to exchange and contraction and obey weakening \cite{pfpl}.
-The natural number $n$ is used as a technical device in our semantics to ensure that the forms shown as being indexed by $n$ arise in a controlled manner to prevent ``out of bounds'' issues, as we will discuss; they would have no corresponding concrete syntax so $n$ can be assumed $0$ in user-defined terms.
-
-The dynamic semantics of static terms is defined as a structural operational semantics by a stepping judgement $\sstep{\st}{\argEnv}{\st\moutput}$, a value judgement $\sval{\st}{\argEnv}$ and an error raised judgement $\serr{\st}{\argEnv}$.
-$\argEnv$ ranges over \emph{argument environments}, which we will  return to when considering opcons in Sec.
-\ref{external-terms}.
-The multi-step judgement $\smanystep{\st}{\argEnv}{\st\moutput}$ is the reflexive, transitive closure of the stepping judgement.
-The normalization judgement $\seval{\st}{\argEnv}{\st'}$ is derivable iff $\smanystep{\st}{\argEnv}{\st'}$ and $\sval{\st'}{\argEnv}$.
+Its syntax is shown in Figure \ref{syntax-SL}.
+The portion of the SL covered by the first row of kinds and static terms, some of which are elided, forms a standard functional language consisting of total functions, polymorphic and inductive kinds, and products and sums  \cite{pfpl}.
+It can be seen as a total subset of ML.%and we will assume standard conveniences in examples.
+
+Only three new  kinds are needed for the SL to serve its role: 1) $\kty$, classifying types (Sec.
+\ref{types}); 2) $\kity$, classifying \emph{quoted translational internal types}, used to {compute} type translations in Sec.
+\ref{sec:type-translations}; and (3) $\kitm$, classifying \emph{quoted translational internal terms}, used to {compute} term translations in Sec.
+\ref{sec:introop}.
+The forms $\sana{n}{\st}$ and $\ssyn{n}$ will  allow tycon definitions to request analysis or synthesis of operator arguments, linking the SL with the  EL in Sec.
+\ref{sec:argument-interfaces}.
+
+The kinding judgement takes the form $\sofkX{\st}{\kappa\moutput}$, where $\kDelta$ and $\kGamma$ are analagous to $\iDelta$ and $\iGamma$ and analagous judgements $\kDelta \vdash \kappa$ and $\kDelta \vdash \kGamma$ are  defined.
+% All such contexts in @$\lambda$ are identified up to exchange and contraction and obey weakening \cite{pfpl}.
+The natural number $n$ is simply a bound used to prevent ``out of bounds'' references to arguments. 
+The computational behavior of static terms (i.e.
+the \emph{static dynamics}) is defined by a stepping judgement $\sstep{\st}{\argEnv}{\st\moutput}$, a value judgement $\sval{\st}{\argEnv}$ and an \emph{error raised} judgement $\serr{\st}{\argEnv}$.
+$\argEnv$ ranges over \emph{argument environments}, which we  return to in Sec.
+\ref{sec:argument-interfaces}.
+The multi-step judgement $\smanystep{\st}{\argEnv}{\st\moutput}$ is the reflexive, transitive closure of the stepping judgement and the normalization judgement $\seval{\st}{\argEnv}{\st'}$ is derivable iff $\smanystep{\st}{\argEnv}{\st'}$ and $\sval{\st'}{\argEnv}$.
 %${\st}{\memD}{\memG}{\aCtx}{\st\moutput}{\memD\moutput}{\memG\moutput}$, where $\memD$, $\memG$ and $\aCtx$ are also technical devices that will be described later; they too can be ignored from the perspective of user code.
@@
 
-\section{Types}\label{types}
-
-\begin{figure*}\begin{mathpar}
+\begin{figure*}[t!]\begin{mathpar}
 \small
-\inferrule[k-parr-ty]{
+\inferrule[k-ty-parr]{
     \sofkX{\st}{\kprod{\kty}{\kty}}
@@
 
-\inferrule[k-ext-ty]{
+\inferrule[k-ty-ext]{
     \tcdef{\tc}{\tcsig{\ktyidx}{\chi}}{\theta} \in \Phi\\
@@
 
-\inferrule[k-other-ty]{
-    \sofkX{\sttyidx}{\kappa \times \kity}
-}{
-    \sofkX{\sty{\keyw{other}[m; \kappa]}{\sttyidx}}{\kty}
+\inferrule[k-ty-other]{
+    \keq{\emptyset}{\kappa}\\
+    \sofkX{\st}{\kappa \times \kity}
+}{
+    \sofkX{\sty{\keyw{other}[m; \kappa]}{\st}}{\kty}
 }\vspace{-10px}
 \end{mathpar}
-\caption{Kinding rules for types, which take the form $\sty{c}{\sttyidx}$ where $c$ is a tycon and $\sttyidx$ is the type index.}
+\caption{Kinding rules for types, which take the form $\sty{c}{\st}$ where $c$ is a tycon and $\st$ is the type index.}
 \label{fig:types}\vspace{-10px}
 \end{figure*}
-
-\noindent
-External types, or simply \emph{types}, are static values of kind $\kty$.
-We write $\istype{\st}{\Phi}$  iff $\sofkn{\emptyset}{\emptyset}{\Phi}{0}{\st}{\kty}$ and $\sval{\st}{\cdot;\emptyset;\Phi}$.
-The introductory form for kind $\kty$ is $\sty{c}{\st}$, where $c$ is a \emph{tycon} and $\st$ is the \emph{type index}.
-The dynamics are simple: the index is eagerly normalized and errors propagate  (the supplement gives the full set of rules for this paper).
-
-The syntax  for tycons given in Figure \ref{syntax-TC} specifies that $c$ can have one of three forms, so there are three corresponding kinding rules for types, shown in Figure \ref{fig:types}.
-
-\paragraph{Function Types} The rule \rulename{k-parr-ty} specifies that the type index of partial function types must be a pair of types.
-We thus say that $\rightharpoonup$ has \emph{index kind} $\kty \times \kty$.
-In practice, we would introduce a desugaring from $\st_1 \rightharpoonup \st_2$ to $\sty{\rightharpoonup}{(\st_1, \st_2)}$.
-%Here, we will benefit by treating it uniformly.
-
-\paragraph{Extension Types} For types constructed by an \emph{extension tycon}, written in small caps, $\tc$, the rule \rulename{k-ext-ty} extracts the index kind of $\tc$ from the tycon context, $\Phi$.
+\section{Types}\label{types}
 
@@
 \end{array}$
-\caption{Syntax of tycons. Metavariables $\tc$ and $\opname{op}$ range over extension tycon and opcon names, respectively, and $m$ ranges over natural numbers.}
+\caption{Syntax of tycons. Metavariables $\tc$, $\opname{op}$ and $m$ range over extension tycon and opcon names and natural numbers, respectively. We omit leading $\cdot$ in examples.}
 \label{syntax-TC}\vspace{-8px}
@@
 
-For example,  $\tcvar{rstr}$ specifies the index kind $\krx$, which classifies static regular expression patterns (defined as an inductive sum kind in the usual way).
-Types constructed by $\tcvar{rstr}$ will classify terms that behave as \emph{regular strings}, i.e.
-strings statically known to be in the specified regular language \cite{sanitation-psp14}.
-For example, $\stx{title} := \sty{\tcvar{rstr}}{\concrx{.+}}$, which was shown in Sec.
-\ref{intro}, will classify non-empty strings and $\stx{conf} := \sty{\tcvar{rstr}}{\concrx{[A-Z]+ \digit\digit\digit\digit}}$ will classify strings having the format of a conference name.
-The type indices are  here written using standard concrete syntax for concision; recent work has specified how to define type-specific (or here, kind-specific) syntax like this composably \cite{TSLs}.
+\noindent
+Types are static values of kind $\kty$, i.e.
+we write $\istype{\st}{\Phi}$  iff $\sval{\st}{\argEnv}$ and $\sofkn{\emptyset}{\emptyset}{\Phi}{n}{\st}{\kty}$.
+All types are of the form $\sty{c}{\st}$, where $c$ is a \emph{tycon} and $\st$ is the \emph{type index}.
+Three kinding rules govern this form, shown in Figure \ref{fig:types}, one for each of the three forms for tycons given in Figure \ref{syntax-TC}.
+The dynamics are simple and tycon-independent: the index is eagerly normalized and errors propagate (see supplement for the complete rules, lemmas and proofs from this paper).
+
+\paragraph{Function Types} In our example, we assumed a desugaring from $\st_1 \rightharpoonup \st_2$ to $\sty{\rightharpoonup}{(\st_1, \st_2)}$.
+The rule \rulename{k-ty-parr} specifies that the type index of partial function types must be a pair of types.
+We thus say that $\rightharpoonup$ has \emph{index kind} $\kty \times \kty$.
+%Here, we will benefit by treating it uniformly.
+
+\paragraph{Extension Types}
+
+For types constructed by an \emph{extension tycon}, written $\tc$, rule (k-ty-ext) checks for a \emph{tycon definition} for $\tc$ in the tycon context, $\Phi$.
+The syntax of tycon contexts is shown in Figure \ref{syntax-TC} and our main examples are in Figure \ref{fig:example-tycons}.
+For now, the only relevant detail is that each tycon defines a \emph{tycon signature}, $\psi$, which in turn defines the index kind used in the second premise of (k-ty-ext).
+
+Types constructed by $\tcvar{rstr}$, e.g.
+$\stx{title}$ and $\stx{conf}$, specify regular expressions as their indices.
+The tycon signature of $\tcvar{rstr}$, $\psi_\text{rstr}$ in Figure \ref{fig:example-tycons}, thus specifies index kind $\krx$, which classifies static regular expression patterns (defined as an inductive sum kind in the usual way).
+We wrote the type indices in our example assuming a standard concrete syntax.
+Recent work has shown how to define such type-specific, or here kind-specific, syntax composably \cite{TSLs}.
 %We define the tycon context containing only the definition of $\tcvar{rstr}$, $\Phi_\text{rstr} := \tcdef{\tcvar{rstr}}{\tcsig{\krx}{\chi_\text{rstr}}}{\theta_\text{rstr}}$.
 
-Our second example is $\tcvar{lprod}$, which defines a variant of labeled product type (labeled products are like record types, but maintain a row  ordering; records are definable in a manner discussed in Sec.
-\ref{sec:introop}, but an ordering simplifies our discussion).
-We choose the index kind of $\tcvar{lprod}$ to be $\klist{\klbl \times \kty}$, where list kinds are defined as inductive sums in the usual way, and $\klbl$ classifies static representations of row labels.
-In a tycon context defining both tycon definitions, we can define a type classifying conference papers, $\stx{paper} := \sty{\tcvar{lprod}}{\{\conclbl{title} : \stx{title}, \conclbl{conf} : \stx{conf}\}}$.
+Under the composed context $\Phi_\text{rstr}\Phi_\text{lprod}$, we defined the labeled product type $\stx{paper}$.
+The index kind of $\tcvar{lprod}$, given by $\psi_\text{lprod}$ in Figure \ref{fig:example-tycons}, is $\klist{\klbl \times \kty}$, where list kinds are defined in the usual way, and $\klbl$ classifies static representations of row labels, and
 %Note that $\istype{\stx{paper}}{\Phi_\text{rstr}\Phi_\text{lprod}}$ and 
-We again use kind-specific syntax, in this case for $\klist{\klbl \times \kty}$.
-
-Two tycon contexts, each defining just one of the tycons just mentioned, are shown in Figure \ref{fig:example-tycons}.\footnote{In examples, we omit leading $\emptyset$, used as the base case for finite mappings, and $\cdot$, used as the base case for finite sequences, for concision.}
-The syntax of tycon contexts in Figure \ref{syntax-TC} specifies that they are simply lists of tycon definitions, $\tcdef{\tc}{\psi}{\theta}$.
-The tycon structure, $\theta$, contains a \emph{translation schema} (Sec.
-\ref{sec:type-translations}) and opcon definitions in an \emph{opcon structure}, $\omega$ (Sec.
-\ref{external-terms}).
-\emph{Tycon signatures}, $\psi$,  have the form $\tcsig{\ktyidx}{\chi}$, where $\ktyidx$ is the tycon's index kind and $\chi$ is the \emph{opcon signature}, defining index kinds for the opcons in the opcon structure.
-The rule (k-ext-ty) only needs the tycon index kind.
-
-The tycon context well-definedness judgement, written $\vdash \Phi$, requires that all tycon names are unique and performs three additional checks, shown in Figure \ref{fig:tycon-context-well-definedness} and described below (we omit the trivial base case, (tcc-emp)).
-We write tycon context concatenation as, e.g., $\Phi_\text{rstr}\Phi_\text{lprod}$.
+we again used kind-specific syntax to approximate a conventional syntax for row specifications.
 
 \paragraph{Other Types}
-The rule (k-otherty) governs types constructed by a tycon of the form $\keyw{other}[m; \kappa]$.
-These will serve only as technical devices to stand in for types other than those in a given tycon context in Sec.
-\ref{metatheory}.
-The index must pair a term of kind $\kappa$  with a static term of kind $\kity$, discussed in Sec.
-\ref{sec:type-translations}.
-These can be thought of as a technical realization of the informal practice of including a ``catch-all'' or token base type (e.g.
+Rule (k-ty-other) governs types constructed by a tycon of the form $\keyw{other}[m; \kappa]$.
+These will serve only as technical devices to stand in for tycons other than those in a given tycon context in Theorem \ref{thm:conservativity}.
+The natural number $m$ serves to ensure there are arbitrarily many of these tycons.
+The index pairs any value of \emph{equality kind} $\kappa$, discussed in Sec.
+\ref{sec:type-equivalence}, with a value of kind $\kity$, discussed in Sec.
+\ref{sec:type-translations}.% These can be thought of as a technical realization of the informal practice of including a ``catch-all'' or token base type (e.g.
 unit or nat) in a calculus.
-%The natural number $m$ serves to create arbitrarily many of these tycons.
-
-\subsection{Type Case Analysis}
 
 \begin{figure}\small
-\hspace{-10px}\begin{tabular}{l|l}
+$\psi_\text{rstr} := \tcsig{\krx}{\introsig{\keyw{Str}}; \opname{conc}[\kunit]; \opname{case}[\keyw{StrPattern}]; ...}$\\
+$\psi_\text{lprod} := \tcsig{\klist{\klbl\times\kty}}{\introsig{\klist{\klbl}}; \opname{\#}[\klbl]; \opname{conc}[\kunit]; ...}$\\
+
+\vspace{-5px}\hspace{-10px}\begin{tabular}{l|ll}
 $\begin{array}{l}\Phi_\text{rstr} := \tcdef{\tcvar{rstr}}{\psi_\text{rstr}}{\\
@@
 \quad\keyw{syn}~\opname{conc} = \stx{lprod/conc};\\
-\quad\cdots}\end{array}$
-\end{tabular}\vspace{3px}\\
-$\psi_\text{rstr} := \tcsig{\krx}{\introsig{\keyw{Str}}; \opname{conc}[\kunit]; \opname{case}[\keyw{StrPattern}]; ...}$\\
-$\psi_\text{lprod} := \tcsig{\klist{\klbl\times\kty}}{\introsig{\klist{\klbl}}; \opname{\#}[\klbl]; \opname{conc}[\kunit]; ...}$
-\caption{Example tycon definitions and signatures.}
+\quad\cdots}\end{array}$ & $\begin{array}{l}
+~\\
+\hspace{-12px}\text{\color{gray} (Sec 3.4)}\\
+\hspace{-12px}\text{\color{gray} (Sec 4.1)}\\
+\hspace{-12px}\text{\color{gray} (Sec 4.4)}\\
+\hspace{-12px}\text{\color{gray} (Sec 4.4)}\\
+~\end{array}
+$
+\end{tabular}\vspace{0px}
+\caption{Example tycon signatures and definitions.}
 \label{fig:example-tycons}\vspace{-8px}
 \end{figure}
+
+\subsection{Type Case Analysis}
+
 \noindent
-Types here can be thought of as arising from a distinguished ``open datatype'' \cite{conf/ppdp/LohH06} defined by the tycon context.
+Types might be thought of as arising from a distinguished ``open datatype'' \cite{conf/ppdp/LohH06} defined by the tycon context.
 Consistent with this view, a type $\st$ can be case analyzed using $\stycase{c}{\st}{\sx}{\st_1}{\st_2}$.
-If the value of $\st$ is constructed by $c$, its type index is bound to $\sx$ and the branch $\st_1$ is taken.
-For totality, a default branch, $\st_2$, must also be provided.
+If the value of $\st$ is constructed by $c$, its type index is bound to $\sx$ and branch $\st_1$ is taken.
+For totality, a default branch, $\st_2$, must be provided.
 For example, the kinding rule for extension tycons is:
@@
 \small
-\inferrule[k-tccase]{
+\inferrule[k-tycase-ext]{
     \sofkX{\st}{\kty}\\
@@
 % \end{mathpar}
-The rule for $c{=}{\rightharpoonup}$ is analagous, but, importantly, no rule for case analyzing against $c{=}\keyw{other}[m; \kappa]$ is defined (these types thus always take the default branch).
+We will see an example of its use in an opcon definition in Sec.
+\ref{sec:targops}.
+The rule for $c{=}{\rightharpoonup}$ is analagous, but, importantly, no rule for $c{=}\keyw{other}[m; \kappa]$ is defined (these types always take the default branch and their indices cannot be examined).
 %The dynamics (see supplement) are straightforwardly consistent with these intuitions.
-We will see an example of its use in an opcon definition in Sec. \ref{sec:targops}.
+
+\subsection{Tycon Contexts}
+% \noindent Figure \ref{syntax-TC} specifies that tycon contexts are simply lists of tycon definitions, $\tcdef{\tc}{\psi}{\theta}$.
+Two tycon contexts, each defining just one of the tycons just mentioned, are shown in Figure \ref{fig:example-tycons}.
+%  Each defines a \emph{tycon structure}, $\theta$, and a \emph{tycon signature}, $\psi$.
+{Tycon signatures}  have the form $\tcsig{\ktyidx}{\chi}$, where $\ktyidx$ is the tycon's index kind and $\chi$ is the \emph{opcon signature}.
+The rules (k-ty-ext) and (k-tycase-ext) only needed the tycon index kind.
+The tycon structure contains a \emph{translation schema} (controlling type translations, Sec.
+\ref{sec:type-translations}) and an \emph{opcon structure}, $\omega$, giving definitions for each opcon in the opcon signature, $\chi$ (controlling term translations, Sec.
+\ref{external-terms}).
+
+\noindent The tycon context well-definedness judgement, $\vdash \Phi$ is given in Figure \ref{fig:tycon-context-well-definedness} (omitting the trivial rule for $\Phi=\cdot$).
+
+\subsection{Type Equivalence}\label{sec:type-equivalence}
+\noindent The first of the three checks in (tcc-ext), and the check in (k-ty-other),  simplifies type equivalence: type index kinds must be \emph{equality kinds}, i.e.
+those for which semantically equivalent values are syntactically equal.
+Equality kinds are defined by the judgement $\keq{\kDelta}{\kappa}$ (see supplement) and are exactly analagous to equality types as in Standard ML \cite{Tofte:89:TheDefinitionOfStandardML}.
+Arrow kinds are not equality kinds.
+
+% \begin{figure}\hfill \fbox{$\vdash \Phi$}\vspace{-25px}
+% \caption{Tycon context well-definedness. We omit the trivial case for when $\Phi=\cdot$ for concision.}
+% \label{fig:tycon-ctxs}
+% \vspace{-8px}
+% \end{figure}
+\subsection{Type Translations}\label{sec:type-translations}
+
+\noindent 
+Recall that a type, $\st$, defines a translation, $\tau$.
+Extension tycons {compute} translations for the types they construct as a function of each type's index by specifying a \emph{translation schema} in the {tycon structure}, $\theta$.
+The translation schema must have kind $\ktyidx \rightarrow \kity$, checked by the third premise of (tcc-ext).
+Terms of kind $\kity$ are introduced by a \emph{quotation form}, $\sqity{\qity}$, where $\qity$ is a \emph{translational internal type}.
+Each form of internal type, $\tau$,  has a corresponding form of translational internal type.
+For example, regular string types have type translations abbreviated $\keyw{str}$.
+Abbreviating the corresponding translational internal type $\hat{\keyw{str}}$, we define the translation schema of $\tcvar{rstr}$ as $\stx{rstr/trans} := \slam{\krx}{\svar{tyidx}}{\sqity{\hat{\keyw{str}}}}$.
+
+%The operational semantics for shared forms are also straightforwardly recursive (see supplemental material).
+
+The syntax for $\qity$ also includes an ``unquote'' form,  $\qtuq{\st}$, so that they can be constructed compositionally, as well as  a form, $\srep{\st}$, that refers to another type's translation.
+These have the following simple kinding rules: \begin{mathpar}\small
+\inferrule[k-ity-unquote]{
+    \sofkX{\st}{\kity}   
+}{
+    \sofkX{\sqity{\qtuq{\st}}}{\kity}
+}
+
+\inferrule[k-ity-trans]{
+    \sofkX{\st}{\kty}
+}{
+    \sofkX{\sqity{\srep{\st}}}{\kity}
+}
+\end{mathpar}
+The semantics for the shared forms propagates the quotation marker recursively to ensure these are reached, e.g.
+\begin{mathpar}
+\small
+\inferrule[k-ity-prod]{
+    \sofkX{\sqity{\qity_1}}{\kity}\\
+    \sofkX{\sqity{\qity_2}}{\kity}
+}{
+    \sofkX{\sqity{\qity_1 \times \qity_2}}{\kity}
+}
+\end{mathpar}
+These are needed in the translation schema for $\tcvar{lprod}$, which generates nested binary product types by folding over the type index and referring to the translations of the types therein.
+We assume the standard $\svar{fold}$ function in defining:
+\[\small
+\begin{array}{l}
+\stx{lprod/trans} := \slam{\klist{\klbl \times \kty}}{\svar{tyidx}}{\svar{fold}~\svar{tyidx}~\sqity{\iunit}~\\
+  \quad (\lambda \svar{h}{:}\klbl \times \kty.\lambda \svar{r}{:}\kity.\sqity{\srep{{\ssnd{\svar{h}}}}\times\qtuq{\svar{r}}}}
+\end{array}\]
+%We assume a standard $\small\svar{fold} :: \kforall{\kalpha_1}{\kforall{\kalpha_2}{\klist{\kalpha_1}\rightarrow\kalpha_2\rightarrow(\kalpha_1\rightarrow\kalpha_2\rightarrow\kalpha_2)\rightarrow\kalpha_2}}$ in defining this translation schema.
 
@@
     \vdash \Phi\\
-    \tc \notin \text{dom}(\Phi)\\
     \keq{\emptyset}{\ktyidx}\\
@@
 \label{fig:tycon-context-well-definedness}
-\vspace{-8px}
+\vspace{-5px}
 \end{figure}
+
+Applying this translation schema to the index of $\stx{paper}$, for example, produces the value  $\stx{paper/trans} := \sqity{\qity_\text{paper/trans}}$ where $\small\qity_\text{paper/trans} := \srep{\stx{title}}\times(\srep{\stx{conf}}\times \iunit)$.
+Note  that references to translations of types are retained in values of kind $\kity$, while unquote forms are eliminated (see supplement for the full semantics of quotations).%It may be helpful to derive that $\stx{paper/trans} :: \kity$ by the rules above.%For simplicity, and to make the point that the choice of representation has only tycon-local implications, our translation schema does not optimize away
+
 \noindent
-% \begin{figure}\hfill \fbox{$\vdash \Phi$}\vspace{-25px}
-% \caption{Tycon context well-definedness. We omit the trivial case for when $\Phi=\cdot$ for concision.}
-% \label{fig:tycon-ctxs}
-% \vspace{-8px}
-% \end{figure}
-
-\subsection{Type Equivalence}\label{sec:type-equivalence}
-\noindent The first check in (tcc-ext) simplifies type equivalence: type index kinds must be \emph{equality kinds}, i.e.
-those for which semantically equivalent values are syntactically equal.
-Equality kinds are defined by the judgement $\keq{\kDelta}{\kappa}$ (see supplement).
-Arrow kinds are not equality kinds, so type indices cannot contain static functions.
-Equality kinds are exactly analagous to equality types as in Standard ML \cite{Tofte:89:TheDefinitionOfStandardML}.
-
-\subsection{Type Translations}\label{sec:type-translations}
-\noindent 
-Recall that every type $\st$ has a type translation, $\tau$.
-Extension tycons {compute} translations for the types they construct as a function of each type's index by specifying a \emph{translation schema} in the {tycon structure}, $\theta$.
-A tycon with index kind $\ktyidx$ must define a translation schema of kind $\ktyidx \rightarrow \kity$, checked by (tcc-ext).
-
-Terms of kind $\kity$ are introduced by a quotation form, $\sqity{\qity}$, where $\qity$ is a \emph{translational internal type}.
-Each form of internal type, $\tau$,  corresponds to a form of translational internal type, $\qity$.
-For example,  $\tcvar{rstr}$ ignores the type index and translate regular strings to strings, of internal type abbreviated $\keyw{str}$.
-Abbreviating the corresponding translational internal type $\hat{\keyw{str}}$, we define $\stx{rstr/trans} := \slam{\krx}{\svar{tyidx}}{\sqity{\hat{\keyw{str}}}}$.
-The kinding and dynamics for these shared forms is simply recursive (maintaining the quotation marker), e.g.
-\begin{mathpar}
-\small
-\inferrule[k-ity-prod]{
-    \sofkX{\sqity{\qity_1}}{\kity}\\
-    \sofkX{\sqity{\qity_2}}{\kity}
-}{
-    \sofkX{\sqity{\qity_1 \times \qity_2}}{\kity}
-}
-\end{mathpar}
-%The operational semantics for shared forms are also straightforwardly recursive (see supplemental material).
-
-The syntax for $\qity$ further includes an ``unquote'' form,  $\qtuq{\st}$, so that they can be constructed compositionally, as well as a form, $\srep{\st}$, that allows one type's translation to refer to another's: \begin{mathpar}\small
-\inferrule[k-ity-unquote]{
-    \sofkX{\st}{\kity}   
-}{
-    \sofkX{\sqity{\qtuq{\st}}}{\kity}
-}
-
-\inferrule[k-ity-trans]{
-    \sofkX{\st}{\kty}
-}{
-    \sofkX{\sqity{\srep{\st}}}{\kity}
-}
-\end{mathpar}
-The unquote form is eliminated during normalization, while references to type translations are retained in values:
-\begin{mathpar}
-\small
-\inferrule[s-ity-unquote-elim]{
-    \sval{\sqity{\qity}}{\argEnv}
-}{
-    \sstep{\sqity{\qtuq{\sqity{\qity}}}}{\argEnv}{\sqity{\qity}}
-}
-
-\inferrule[s-ity-trans-val]{
-    \sval{\st}{\argEnv}
-}{
-    \sval{\sqity{\srep{\st}}}{\argEnv}
-}
-\end{mathpar}
-These forms are needed in the translation schema for $\tcvar{lprod}$, which generates nested binary product types (though we could also have used, e.g., an internal list) by folding over the type index and referring to the translations of the types therein.
-$\stx{lprod/trans}:=$
-\[\small
-\begin{array}{l}
-\slam{\klist{\klbl \times \kty}}{\svar{tyidx}}{\svar{fold}~\svar{tyidx}~\sqity{\iunit}~\\
-  \quad (\lambda \svar{h}{:}\klbl \times \kty.\lambda \svar{r}{:}\kity.\sqity{\srep{{\ssnd{\svar{h}}}}\times\qtuq{\svar{r}}}}
-\end{array}\]
-%We assume a standard $\small\svar{fold} :: \kforall{\kalpha_1}{\kforall{\kalpha_2}{\klist{\kalpha_1}\rightarrow\kalpha_2\rightarrow(\kalpha_1\rightarrow\kalpha_2\rightarrow\kalpha_2)\rightarrow\kalpha_2}}$ in defining this translation schema.
-
-Applying this translation schema to the index of $\stx{paper}$, for example, produces  $\stx{paper/trans} := \sqity{\qity_\text{paper/trans}}$ where $\small\qity_\text{paper/trans} := \srep{\stx{title}}\times(\srep{\stx{conf}}\times \iunit)$.
-Note that our schema did not remove the trailing unit type for simplicity.% (and to again emphasize that this choice is local).
-%It may be helpful to derive that $\stx{paper/trans} :: \kity$ by the rules above.%For simplicity, and to make the point that the choice of representation has only tycon-local implications, our translation schema does not optimize away
-
 \subsubsection{Selective Type Translation Abstraction}\label{sec:selective-type-translation-abstraction}
 \noindent
-References to type translations are maintained in values of kind $\kity$ like this to  allow us to selectively hold them abstract, which will be the key to translation independence.
+References to type translations are maintained in values like this to  allow us to selectively hold them abstract.
 This can be thought of as analagous to the process in ML by which the true identity of an abstract type in a module is held abstract outside the module until after typechecking.
-The judgement $\small\tdeabs{\Phi}{c}{\qity}{\memD}{\ity\moutput}{\memD\moutput}$ relates a normalized translational internal type $\qity$ to an internal type $\tau$, called a \emph{selectively abstracted type translation} because references to translations of types constructed by a tycon other than the \emph{delegated tycon}, $c$, are replaced by a corresponding type variable, $\alpha$.
-The \emph{type translation store} $\memD ::= \emptyset ~|~ \memD, \st \leftrightarrow \ity/\alpha$ maintains this correspondence between types, their actual translations and the distinct type variables which appear in their place, e.g.
-$\small\tdeabs{\Phi_\text{rstr}\Phi_\text{lprod}}{\tcvar{lprod}}{\qity_\text{paper/trans}}{\emptyset}{\tau_\text{paper/abs}}{\memD_\text{paper/abs}}$
-where $\small\tau_\text{paper/abs} := \alpha_1 \times (\alpha_2 \times \iunit)$ and 
+The judgement $\small\tdeabs{\Phi}{c}{\qity}{\memD}{\ity\moutput}{\memD\moutput}$ relates a normalized translational internal type $\qity$ to an internal type $\tau$, called a \emph{selectively abstracted type translation} because references to translations of types \emph{constructed by a tycon other than the delegated tycon}, $c$, are replaced by a corresponding type variable, $\alpha$.
+For example, $\small\tdeabs{\Phi_\text{rstr}\Phi_\text{lprod}}{\tcvar{lprod}}{\qity_\text{paper/trans}}{\emptyset}{\tau_\text{paper/abs}}{\memD_\text{paper/abs}}$
+where $\small\tau_\text{paper/abs} := \alpha_1 \times (\alpha_2 \times \iunit)$.
+
+The \emph{type translation store} $\memD ::= \emptyset ~|~ \memD, \st \leftrightarrow \ity/\alpha$ maintains the correspondence between types, their actual translations and the distinct type variables appearing in their place,  e.g.
 $\memD_\text{paper/abs}  := \stx{title} \leftrightarrow \keyw{str}/\alpha_1, \stx{conf} \leftrightarrow \keyw{str}/\alpha_2$.
-
-%This process of selectively holding the translations of all types constructed by any tycon other than the ``delegated tycon'' abstract will be critical for conservativity to hold below.
-Here, the translation of the type is being held abstract, rather than the type itself, and type translations are computed from indices, rather than declared ``literally''.
-
 The judgement $\memD \leadsto \delta\moutput : \Delta\moutput$ constructs the $n$-ary \emph{type substitution}, $\delta ::= \emptyset ~|~ \delta, \tau/\alpha$, and corresponding internal type formation context, $\Delta$, implied by the type translation store $\memD$.
 For example, $\memD_\text{paper/abs} \leadsto \delta_\text{paper/abs} : \Delta_\text{paper/abs}$ where $\delta_\text{paper/abs} := \keyw{str}/\alpha_1, \keyw{str}/\alpha_2$ and $\Delta_\text{paper/abs} := \alpha_1, \alpha_2$.
+
 We can apply type substitutions to internal types, terms and typing contexts, written $[\delta]\ity$, $[\delta]\iota$ and $[\delta]\Gamma$, respectively.
@@
 the actual type translation of $\stx{paper}$.
-Indeed, we can now give the rule defining the type translation judgement, $\vdash_\Phi \st \leadsto \tau$, mentioned in Sec.
+Indeed, we can now define the type translation judgement, $\vdash_\Phi \st ~\mathtt{type} \leadsto \tau$, mentioned in Sec.
 \ref{overview}.
-We simply determine any selectively abstract translation, then apply the induced substitution:
+We simply determine any selectively abstracted translation, then apply the implied substitution:
 \begin{mathpar}\small
@@
 }{
-    \vdash_\Phi \st \leadsto [\delta]\tau
+    \vdash_\Phi \st ~\mathtt{type} \leadsto [\delta]\tau
 }
@@
 The rules for the selective type translation abstraction judgement recurse generically over shared forms in $\qity$. Only sub-terms of form $\srep{\st}$ are interesting. 
-The translation of an extension type  is determined by calling the translation schema and checking that the type translation it generates is closed except for type variables tracked by $\memD'$.
+The translation of an extension type  is determined by calling the translation schema and validating that the type translation it generates is closed except for type variables tracked by $\memD'$.
 If constructed by the delegated tycon, it is not held abstract:
@@
 \small
-\inferrule[abs-tc-delegated]{
+\inferrule[abs-ext-delegated]{
     \tcdef{\tc}{\psi}{\tcstruct{\stx{schema}}{\omega}} \in \Phi\\
@@
 \end{mathpar}
-Otherwise, it is held abstract via a fresh type variable added to the store (the supplement has the rule (abs-ty-stored) for retrieving it if already there):
+Otherwise, it is held abstract via a fresh type variable added to the store (the supplement gives rule (abs-ty-stored) for retrieving it if already there):
 \begin{mathpar}
 \small
-\inferrule[abs-tc-not-delegated-new]{
+\inferrule[abs-ext-not-delegated-new]{
     c \neq \tc\\
@@
     \seval{\stx{schema}(\sttyidx)}{\cdot;\emptyset;\Phi}{\sqity{\qity}}\\
-    \tdeabs{\Phi}{\tc}{\qity}{\memD}{\ity}{\memD'}\\\\
+    \tdeabs{\Phi}{c}{\qity}{\memD}{\ity}{\memD'}\\\\
     \memD' \leadsto \delta : \Delta\\
@@
 \end{mathpar}
-The translation of an ``other'' type is given in its index:
+The translations of ``other'' types are given directly in their indices (in Sec.
+\ref{metatheory}, we will replace some extension types with other types, and this is how we  preserve their translations):
 \begin{mathpar}
@@
 \inferrule[abs-other-delegated]{
-    \tdeabs{\Phi}{\keyw{other}[m;\kappa]}{\qity}{\memD}{\tau}{\memD'}
+    \tdeabs{\Phi}{\keyw{other}[m;\kappa]}{\qity}{\memD}{\tau}{\memD'}\\
+    \memD' \leadsto \delta : \Delta\\
+    \Delta \vdash \tau\\
 }{
@@
 \end{mathpar}
-Rule (abs-other-not-delegated-new) is in the supplement.
-
-The translations of function types are not held abstract, so that lambdas can be the sole binding construct in the EL:
-\begin{mathpar}
-\small
-\inferrule[abs-parr]{
-    \tdeabs{\Phi}{c}{\srep{\st_1}}{\memD}{\ity_1}{\memD'}\\
-    \tdeabs{\Phi}{c}{\srep{\st_2}}{\memD'}{\ity_2}{\memD''}
-}{
-    \tdeabs{\Phi}{c}{\srep{\sty{\rightharpoonup}{(\st_1, \st_2)}}}{\memD}{\tau_1 \rightharpoonup \tau_2}{\memD''}
-}
-\end{mathpar}
-
-% \paragraph{Summary}
-% In summary, types (e.g.
-$\stx{paper}$) are constructed by applying tycons to type indices.
-Extension tycons compute quoted type translations ($\stx{paper/trans}$), which then can determine many selectively abstracted type translations depending on the delegated tycon (e.g.
-$\tau_\text{paper/abs}$).
-From any such, the actual type translation is determined ($\tau_\text{paper}$).
-
-%\vspace{-5px}
- %We have that external typing contexts obey the standard structural congruences: weakening, exchange and contraction.
-\section{External Terms}\label{external-terms}
-\noindent 
-Now that we have established how types are constructed and how selectively abstracted and from there actual type translations are computed by static functions, we are ready to give the semantics for external terms, shown in Figure \ref{typing}.
-
-Because we are defining a bidirectional type system, a subsumption rule is needed to allow synthetic terms to be analyzed against an equivalent type.
-Per Sec.
-\ref{sec:type-equivalence}, equivalent types must be  syntactically identical at normal form, and we consider analysis only if $\istype{\st}{\Phi}$, so the rule (subsume) is straightforward.
-To use an analytic term in a synthetic position, the programmer must provide a type ascription, written $e : \st$.
-The ascription is kind checked and normalized to a type before being used for analysis by rule (ascribe).
-
-Variables and functions behave in the standard manner given our definitions of types and type translations (used to generate ascriptions in the IL).
-We use Plotkin's fixpoint operator for general recursion (cf.
-\cite{pfpl}), and define both lambdas and fixpoints only analytically for simplicity.
-
-\subsection{Generalized Introductory Operations}\label{sec:introop}
-The translation of the \emph{generalized intro operation}, written $\small\eintro{\sttmidx}{\es}$, is determined by the tycon of the type it is being analyzed against as a function of the type's index, the \emph{term index}, $\sttmidx$, and the \emph{argument list}, $\es$.
-
-Before discussing rules (ana-intro) and (ana-intro-other), we note that we can recover a variety of standard concrete introductory forms by a purely syntactic desugaring to this abstract form (and thus allow their use at more than one type).
-For example, for regular strings we can use the string literal form, $\texttt{"s"}$, which desugars to $\eintro{\texttt{"s"}_\text{SL}}{\cdot}$, i.e.
-the term index is the corresponding static value of kind $\keyw{Str}$, indicated by a subscript for clarity.
-Similarly, for labeled products, records, objects and so on, we can define a generalized labeled collection form, $\{\mathtt{lbl}_1=e_1, \ldots, \mathtt{lbl}_n=e_n\}$, that desugars to $\eintro{\texttt{[}\mathtt{lbl}_1, \ldots, \mathtt{lbl}_n\texttt{]}}{e_1; \ldots; e_n}$, i.e.
-a list constructed from the row labels is the term index and the corresponding row values are the arguments.
-In both cases, the term index captures  static portions of the concrete form and the arguments capture all external sub-terms.
-Additional desugarings are shown in the supplement and a technique based on \cite{TSLs} could be introduced to allow tycon providers to define more such desugarings  composably.
+Rule (abs-other-not-delegated-new) is analagous to (abs-ext-not-delegated-new), and is shown in the supplement.
 
 \begin{figure*}[t]
-\small\fbox{$\eanaX{e}{\st}{\iota}$}~
-\fbox{$\esynX{e}{\st}{\iota}$}\vspace{-3px}
+\small\fbox{$\eanaX{e}{\st}{\iota\moutput}$}~
+\fbox{$\esynX{e}{\st}{\iota\moutput}$}\vspace{-3px}
 \begin{mathpar}
@@
 \small\inferrule[syn-var]{
-  x \Rightarrow \st \in \Upsilon
+  x : \st \in \Upsilon
 }{
@@
 \inferrule[ana-fix]{
-  \eana{\Upsilon, x \Rightarrow \st}{\Phi}{e}{\st}{\iota}\\\\
-  \sfinalrepX{\st}{\tau}
+  \eana{\Upsilon, x : \st}{\Phi}{e}{\st}{\iota}\\\\
+  \tytrans{\Phi}{\st}{\tau}
 }{
@@
 \inferrule[ana-lam]{
-    \eana{\Upsilon, x \Rightarrow \st_1}{\Phi}{e}{\st_2}{\iota}\\
-    \sfinalrepX{\st_1}{\tau_1}
+    \eana{\Upsilon, x : \st_1}{\Phi}{e}{\st_2}{\iota}\\\\
+    \tytransX{\st_1}{\tau_1}
 }{
@@
 
+\inferrule[syn-lam]{
+    \sofkz{\emptyset}{\emptyset}{\Phi}{\st_1}{\kty}\\
+    \st_1 \Downarrow_{\cdot;\emptyset;\Phi} \st_1'\\\\
+    \esyn{\Upsilon, x : \st_1'}{\Phi}{e}{\st_2}{\iota}\\
+    \vdash_\Phi \st_1'~\mathtt{type} \leadsto \tau_1
+}{
+    \esynX{\elam{\st_1}{x}{e}}{\sty{\rightharpoonup}{(\st_1', \st_2)}}{\ilam{\tau_1}{x}{\iota}}
+}
+
 \inferrule[syn-ap]{
-  \esynX{e_1}{\sty{\rightharpoonup}{(\st_1, \st_2)}}{\iota_1}\\
+  \esynX{e_1}{\sty{\rightharpoonup}{(\st_1, \st_2)}}{\iota_1}\\\\
   \eanaX{e_2}{\st_2}{\iota_2}
@@
 \inferrule[ana-intro]{
-  \tcdef{\tc}{\tcsig{\_}{\chi}}{\tcstruct{\_}{\omega}} \in \Phi\\\\
+  \tcdef{\tc}{\tcsig{\ktyidx}{\chi}}{\tcstruct{\stx{schema}}{\omega}} \in \Phi\\\\
   \introsig{\klitidx} \in \chi\\
@@
   \esynX{e_\text{targ}}{\sty{\tc}{\sttyidx}}{\iota_\text{targ}}\\\\
-  \tcdef{\tc}{\tcsig{\_}{\chi}}{\tcstruct{\_}{\omega}} \in \Phi\\\\
+  \tcdef{\tc}{\tcsig{\ktyidx}{\chi}}{\tcstruct{\stx{schema}}{\omega}} \in \Phi\\\\
   \opsig{\opname{op}}{\klitidx} \in \chi\\
@@
 \inferrule[ana-intro-other]{
-    \sofkz{\emptyset}{\emptyset}{\Phi}{\stx{def}}{\kargs \rightarrow \kitm}\\\\
     |\es| = n\\
-    \keyw{args}(n)=\stx{args}\\
-    \stx{def}(\stx{args}) \Downarrow_{\es; \Upsilon; \Phi} \sqitm{\qitm}\\\\
+    \sofk{\emptyset}{\emptyset}{\Phi}{\sqitm{\qitm}}{\kitm}\\
+    \sval{\sqitm{\qitm}}{\es;\Upsilon;\Phi}\\\\
     \trvalidate{\es;\Upsilon;\Phi}{\keyw{other}[m;\kappa]}{\qitm}{{\sty{\keyw{other}[m;\kappa]}{\sttyidx}}}{\iota}
 }{
-    \eanaX{\eintro{\stx{def}}{\es}}{\sty{\keyw{other}[m;\kappa]}{\sttyidx}}{\iota}
-}
-~~~~~~
+    \eanaX{\eintro{\sqitm{\qitm}}{\es}}{\sty{\keyw{other}[m;\kappa]}{\sttyidx}}{\iota}
+}
+
 \inferrule[syn-targ-other]{
 \esynX{e_\text{targ}}{\sty{\keyw{other}[m;\kappa]}{\sttyidx}}{\iota_\text{targ}}\\\\
-  \sofkz{\emptyset}{\emptyset}{\Phi}{\stx{def}}{\kargs\rightarrow(\kty\times\kitm)}\\\\
     |e_\text{targ}; \es| = n\\
-    \keyw{args}(n)=\stx{args}\\
-  \stx{def}(\stx{args}) \Downarrow_{(e_\text{targ}; \es);\Upsilon;\Phi} (\st, \sqitm{\qitm})\\\\
-  \trvalidate{(e_\text{targ}; \es);\Upsilon;\Phi}{\keyw{other}[m; \kappa]}{\qitm}{{\st}}{\iota}
-  %\snorm{\stx{def}~\sttyidx~\stmidx~\stx{args}}{\emptyset}{\memG_0}{{\rightharpoonup}, \tc; \Upsilon; \Phi}{\sqitm{\itm_\text{abs}}}{\memD}{\memG}\\\\
-  %\snorm{\srep{\sty{\tc}{\sttyidx}}}{\memD}{\cdot}{{\rightharpoonup}, \tc;\emptyset;\Phi}{\sqity{\ity_\text{abs}}}{\memD'}{\cdot}\\\\
-}{
-     \esynX{\etarg{\opname{op}}{\stx{def}}{e_\text{targ}}{\es}}{\st}{[\delta][\gamma]\iota_\text{abs}}
+    \sofk{\emptyset}{\emptyset}{\Phi}{\sqitm{\qitm}}{\kitm}\\
+    \sval{\sqitm{\qitm}}{\es;\Upsilon;\Phi}\\\\
+    \istype{\st}{\Phi}\\
+    \trvalidate{(e_\text{targ}; \es);\Upsilon;\Phi}{\keyw{other}[m;\kappa]}{\qitm}{{\st}}{\iota}
+}{
+     \esynX{\etarg{\opname{op}}{(\st, \sqitm{\qitm})}{e_\text{targ}}{\es}}{\st}{\iota}
 }\vspace{-5px}
@@
 \caption{Typing}
-\label{typing}
+\label{typing}\vspace{-8px}
 \end{figure*}
@@
     \vdash_\Phi \omega \sim \tcsig{\ktyidx}{\chi}\\
-    \opname{op}\notin\text{dom}(\chi)\\
+    %\opname{op}\notin\text{dom}(\chi)\\
     \emptyset \vdash \klitidx\\\\
@@
 
-Let us now derive $\small\eana{\Upsilon_\text{ex}}{\Phi_\text{rstr}\Phi_\text{lprod}}{e_\text{ex}}{\stx{paper}}{\iota_\text{ex}}$ where $\small\Upsilon_\text{ex}  := title \Rightarrow \stx{title}$ and $\small e_\text{ex} := \{\conclbl{title}=title, \conclbl{conf}=\texttt{"EXMPL 2015"}\}$.
-The  translation will be $\small\iota_\text{ex} := (title, (\texttt{"EXMPL 2015"}_\text{IL}, ()))$, where $\small\texttt{"EXMPL 2015"}_\text{IL}$ is an internal string (of internal type $\keyw{str}$).
-
-The first premise of (ana-intro) extracts the tycon definition for the tycon of the type the intro form is being analyzed against.
-In this example, this is $\tcvar{lprod}$.
-We will use this as the {delegated tycon} in the final premises of the rule.
-
-The second premise extracts the \emph{intro term index kind}, $\klitidx$, from the \emph{opcon signature}, $\chi$, and the third premise checks the provided term index against this kind.
+The translations of function types are not held abstract, so that lambdas, which are built in, can be the sole binding construct in the EL:
+\begin{mathpar}
+\small
+\inferrule[abs-parr]{
+    \tdeabs{\Phi}{c}{\srep{\st_1}}{\memD}{\ity_1}{\memD'}\\
+    \tdeabs{\Phi}{c}{\srep{\st_2}}{\memD'}{\ity_2}{\memD''}
+}{
+    \tdeabs{\Phi}{c}{\srep{\sty{\rightharpoonup}{(\st_1, \st_2)}}}{\memD}{\tau_1 \rightharpoonup \tau_2}{\memD''}
+}
+\end{mathpar}
+
+% \paragraph{Summary}
+% In summary, types (e.g.
+$\stx{paper}$) are constructed by applying tycons to type indices.
+Extension tycons compute quoted type translations ($\stx{paper/trans}$), which then can determine many selectively abstracted type translations depending on the delegated tycon (e.g.
+$\tau_\text{paper/abs}$).
+From any such, the actual type translation is determined ($\tau_\text{paper}$).
+
+%\vspace{-5px}
+ %We have that external typing contexts obey the standard structural congruences: weakening, exchange and contraction.
+\section{External Terms}\label{external-terms}
+
+\noindent 
+Now that we have established how types are constructed and how type translations are computed, we are ready to give the semantics for external terms, shown in Figure \ref{typing}.
+
+Because we are defining a bidirectional type system, the rule (subsume) is needed to allow synthetic terms to be analyzed against an equivalent type.
+Per Sec.
+\ref{sec:type-equivalence}, equivalent types must be  syntactically identical at normal form, and we consider analysis only if $\istype{\st}{\Phi}$, so the rule is straightforward.
+To use an analytic term in a synthetic position, the programmer must provide a type ascription, written $e : \st$.
+The ascription is kind checked and normalized to a type before being used for analysis by rule (ascribe).
+
+Rules (syn-var) states that variables synthesize types, as is standard.
+Functions operate in the standard manner given our definitions of types and type translations (used to generate annotations in the IL).
+We use Plotkin's fixpoint operator for general recursion (cf.
+\cite{pfpl}), and define it only analytically with rule (ana-fix).
+We also define an analytic form of lambda without a type ascription to emphasize that bidirectional typing allows you to omit type ascriptions in analytic positions.% Rule (syn-ap) is standard.
+
+\subsection{Generalized Intro Operations}\label{sec:introop}
+\noindent The meaning of the \emph{generalized intro operation}, written $\small\eintro{\sttmidx}{\es}$, is determined by the tycon of the type it is being analyzed against as a function of the type's index, the \emph{term index}, $\sttmidx$, and the \emph{argument list}, $\es$.
+
+Before discussing rules (ana-intro) and (ana-intro-other), we note that we can recover a variety of standard concrete introductory forms by desugaring.
+For example, the string literal form, $\texttt{"s"}$, desugars to $\eintro{\texttt{"s"}_\text{SL}}{\cdot}$, i.e.
+the term index is the corresponding static value of kind $\keyw{Str}$ and there are no arguments.
+Similarly, we define a generalized labeled collection form, $\{\mathtt{lbl}_1=e_1, \ldots, \mathtt{lbl}_n=e_n\}$, that desugars to $\eintro{\texttt{[}\mathtt{lbl}_1, \ldots, \mathtt{lbl}_n\texttt{]}}{e_1; \ldots; e_n}$, i.e.
+a static list constructed from the row labels is the term index and the corresponding row values are the arguments.
+Additional desugarings are discussed in the supplement.
+The literal form in $e_\text{paper}$, from Sec.
+\ref{overview}, for example, desugars to $e_\text{paper}:=\eintro{\texttt{[}\conclbl{title}, \conclbl{conf}\texttt{]}}{title; \eintro{\concstr{EXMPL 2015}_\text{SL}}{\cdot}}$.
+%In both cases, the term index captures  static portions of the concrete form and the arguments capture all external sub-terms.
+% (and a technique based on \cite{TSLs} could be introduced to allow tycon providers to define more desugarings  composably).
+
+Let us now derive the typing judgement in Sec.
+\ref{overview}.
+We first apply (syn-lam), which will ask $(e_\text{paper} : \stx{paper})$ to synthesize a type in the typing context $\Upsilon_\text{ex}=title : \stx{title}$.
+Then (ascribe) will analyze $e_\text{paper}$ against $\stx{paper}$ via (ana-intro).
+
+The first premise of (ana-intro) simply finds the tycon definition for the tycon of the type provided for analysis, in this case $\tcvar{lprod}$.
+% We will use this as the {delegated tycon} in the final premises of the rule.
+The second premise extracts the \emph{intro term index kind}, $\klitidx$, from the \emph{opcon signature}, $\chi$, it specifies.
 This is simply the kind of term index expected by the tycon, e.g.
-in Figure \ref{fig:example-tycons}, $\tcvar{lprod}$ specifies $\klist{\klbl}$, so that it can use the labeled collection form, while $\tcvar{rstr}$ specifies an intro index kind of $\keyw{Str}$, so that it can use the string literal form.
+in Figure \ref{fig:example-tycons}, $\tcvar{lprod}$ specifies $\klist{\klbl}$, so that it can use the labeled collection form, while $\tcvar{rstr}$ specifies  $\keyw{Str}$, so that it can use the string literal form.
+The third premise checks the provided term index against this kind.
 %\[\small\begin{array}{lcl}
@@
 The fourth premise extracts the \emph{intro opcon definition} from the \emph{opcon structure}, $\omega$, of the tycon structure, calling it $\stx{def}$.
-This is a static function that is applied, in the seventh premise, to determine whether the term is well-typed, raising an error if not or determining the translation of the term if so.
-The function has access to the type index, the term index and an interface to the list of arguments, and its kind is checked by the  judgement $\vdash_\Phi \omega \sim \psi$, which appeared as the final premise of the rule (tcc-ext) and is defined in Figure \ref{ocstruct}.
-For example, the intro opcon definition for $\tcvar{rstr}$ is $\small\stx{rstr/intro}:=$
-\[\small\begin{array}{l}
-    \slam{\krx}{\svar{tyidx}}{\slam{\kstr}{\svar{tmidx}}{\slam{\kargs}{\svar{args}}{\\
-\quad \keyw{let}~\svar{aok} :: \kunit = \svar{arity0}~\svar{args}~\keyw{in}~\\
-\quad \keyw{let}~\svar{rok} :: \kunit = \svar{rmatch}~\svar{tyidx}~\svar{tmidx}~\keyw{in}~\svar{str\_of\_Str}~\svar{tmidx}
-}}}
-\end{array}\]
-Because regular strings are implemented as strings, this intro opcon definition is straightforward.
-It begins by making sure that no arguments were passed in (we will return to arguments and the fifth and sixth premises of (ana-intro) with the next example), using the helper function $\svar{arity0} :: \kargs \rightarrow \kunit$ defined such that any non-empty list will raise an error, via the static term $\sraise{\kunit}$.
+This is a static function that is applied, in the seventh premise, to determine whether the term is well-typed, raising an error if not or computing a translation if so.
+Its kind is checked by the  judgement $\vdash_\Phi \omega \sim \psi$, which was the final premise of the rule (tcc-ext) and is defined in Figure \ref{ocstruct}.
+Rule (ocstruct-intro) specifies that it has access to the type index, the term index and an interface to the list of arguments, discussed below, and returns a \emph{quoted translational internal term} of kind $\kitm$, analagous to $\kity$.
+The intro opcon definitions for $\tcvar{rstr}$ and $\tcvar{lprod}$ are given in Figures \ref{fig:rstr-intro} and \ref{fig:lprod-intro}.
+
+Though the latter will be encountered first in our example derivation, let us first consider the intro opcon definition in Figure \ref{fig:rstr-intro} because it is more  straightforward.
+It will be used to analyze the row value $\small\eintro{\concstr{EXMPL 2015}_\text{SL}}{\cdot}$ against $\stx{conf}$.
+It begins by making sure that no arguments were passed in using the helper function $\svar{arity0} :: \kargs \rightarrow \kunit$ defined such that any non-empty list will raise an error, via the static term $\sraise{\kunit}$.
 In practice, the tycon provider would specify an error message here.
@@
 
-The introductory form for kind $\kitm$ is also a quotation form, $\sqitm{\qitm}$, where $\qitm$ is a \emph{translational internal term}.
-It is analagous to the form $\sqity{\qity}$ for $\kity$ in Sec.
+\begin{figure}\vspace{-5px}
+$\small\begin{array}{l}
+    \slam{\krx}{\svar{tyidx}}{\slam{\kstr}{\svar{tmidx}}{\slam{\kargs}{\svar{args}}{\\
+\quad \keyw{let}~\svar{aok} :: \kunit = \svar{arity0}~\svar{args}~\keyw{in}~\\
+\quad \keyw{let}~\svar{rok} :: \kunit = \svar{rmatch}~\svar{tyidx}~\svar{tmidx}~\keyw{in}\\
+\quad \svar{str\_of\_Str}~\svar{tmidx}
+}}}
+\end{array}$
+\caption{The intro opcon definition for $\tcvar{rstr}$,  $\stx{rstr/intro}$.}
+\label{fig:rstr-intro}\vspace{-8px}
+\end{figure}
+
+Static terms of kind $\kitm$ are introduceed by the quotation form, $\sqitm{\qitm}$, where $\qitm$ is a \emph{translational internal term}.
+This is analagous to the form $\sqity{\qity}$ for $\kity$ in Sec.
 \ref{sec:type-translations}.
@@
 %The supplement gives the analagous rules.
-The final two forms of translational internal term are $\anatrans{n}{\st}$ and $\syntrans{n}$.
+The two interesting forms of translational internal term are $\anatrans{n}{\st}$ and $\syntrans{n}$.
 These stand in for the translation of argument $n$, the first if it arises via analysis against type $\st$ and the second if it arises via type synthesis.
@@
 
-\begin{figure}\vspace{-10px}
-\[\small
+\begin{figure}\vspace{-8px}
+$\small
 \begin{array}{l}
@@
 \quad\quad\quad \sqitm{(\quq{\svar{rowtr}}, \quq{\svar{r}})
-}\end{array}\]\vspace{-10px}
-\caption{The intro opcon definition for $\tcvar{lprod}$.}
-\label{fig:lprod-intro}\vspace{-8px}
+}\end{array}$
+\caption{The intro opcon definition for $\tcvar{lprod}$, $\stx{lprod/intro}$.}
+\label{fig:lprod-intro}\vspace{-10px}
 \end{figure}
 
-The first line checks that the type provided is inhabited, in this case by checking that there are no duplicate labels via the helper function $\svar{uniqmap} :: \klist{\klbl \times \kty}\rightarrow \kunit$, raising an error if there are.
-An alternative strategy may have been to use an abstract kind that ensured that such type indices could not have been constructed, but to be compatible with our equality kind restriction, this would require support for abstract equality kinds, analagous to abstract equality types in SML.
-We chose not to formalize these for simplicity, and to demonstrate this general technique.
-An analagous technique could be used to implement record types by requiring that the index be sorted (see supplement).%, leaving indices where the labels did not appear sorted uninhabited.
+The first line checks that the type provided is inhabited, in this case by checking that there are no duplicate labels via the helper function $\svar{uniqmap} :: \klist{\klbl \times \kty}\rightarrow \kunit$, raising an error if there are (we briefly discuss alternative strategies in the supplement).
+%An analagous technique could be used to implement record types by requiring that the index be sorted (see supplement).%, leaving indices where the labels did not appear sorted uninhabited.
 In both cases, we could provide a static helper function of kind $\klist{\klbl\times\kty}\rightarrow \kty$ that checked well-formedness immediately before constructing the requested type.
-
-The rest of this opcon definition folds over the three lists provided as input: the list mapping labels to types provided as the type index, the list of labels provided as the term index, and the list of argument interfaces.
+The rest of the definition folds over the three lists provided as input: the list mapping row labels to types provided as the type index, the list of labels provided as the term index, and the list of argument interfaces, which give access to the corresponding row values.
 We assume a straightforward helper function, $\svar{fold3}$, that raises an error if the three lists are not of the same length.
 The base case is the translational empty product.
-The recursive case checks, for each row, that the label provided in the term index matches the label provided in the type index, using a helper function $\svar{lbleq} :: \klbl \rightarrow \klbl \rightarrow \kunit$.
+
+The recursive case checks, for each row, that the label provided in the term index matches the label in the type index using helper function $\svar{lbleq} :: \klbl \rightarrow \klbl \rightarrow \kunit$.
 Then, we request type analysis of the corresponding argument, $\svar{rowarg}$, against the type in the type index, $\svar{rowty}$, by writing $\svar{ana}~\svar{rowarg}~\svar{rowty}$.
-Here, $\svar{ana}$ is a helper function defined below that triggers type analysis of the provided argument.
-If this succeeds, it evaluates to an translational internal term  of the form $\sqitm{\anatrans{n}{\st}}$, where $n$ is the position of $\svar{rowarg}$ in $\svar{args}$ and $\st$ is the value of $\svar{rowty}$.
+Here, $\svar{ana}$ is a helper function defined in Sec.
+\ref{sec:argument-interfaces} below that triggers type analysis of the provided argument.
+If this succeeds, it evaluates to a translational internal term  of the form $\sqitm{\anatrans{n}{\st}}$, where $n$ is the position of $\svar{rowarg}$ in $\svar{args}$ and $\st$ is the value of $\svar{rowty}$.
 If analysis fails, it raises an error.
-The final line constructs a nested tuple based on this translation and the recursive result.
-Taken together, the translational internal term that will be generated for our example involving $e_\text{ex}$ above is $\small\qitm_\text{ex} := (\anatrans{0}{\stx{title}}, (\anatrans{1}{\stx{conf}}, ()))$, i.e.
+The final line constructs a nested tuple based on the row value's translation and the recursive result.
+Taken together, the translational internal term that will be generated for our example involving $e_\text{paper}$ above is $\small\qitm_\text{paper} := (\anatrans{0}{\stx{title}}, (\anatrans{1}{\stx{conf}}, ()))$, i.e.
 it simply recalls that the two arguments were analyzed against $\stx{title}$ and  $\stx{conf}$, without yet inserting their translations directly.
-This is done by the \emph{translation validation judgement}, which occurs in the final premise. We describe argument interfaces and translation validation next.
-
-\subsection{Argument Interfaces} 
+This will be done after \emph{translation validation}, triggered by the final premise of (ana-intro) and described in Sec.
+\ref{sec:translation-validation}.
+%We describe argument interfaces and translation validation next.
+
+\subsection{Argument Interfaces} \label{sec:argument-interfaces}
 \noindent
@@
 $\svar{ana} := \slam{\karg}{\svar{arg}}{\keyw{fst}(\svar{arg})}$.
-To actually perform analysis or synthesis, we must provide a link between the dynamics of the static language and the typing rules.
+To actually perform analysis or synthesis, we must provide a link between the dynamics of the static language and the EL's typing rules.
 This is purpose of the static forms $\sana{n}{\st}$ and $\ssyn{n}$.
-We provide each opcon definition with an \emph{argument interface list}, which  a static list of length $n$ where the $i$th entry is $(\slam{\kty}{\svar{ty}}{\sana{i}{\svar{ty}}}, \slam{\kunit}{\_}{\ssyn{i}})$.
-It is generated by the judgement $\keyw{args}(n)=\stx{args}$, where $n$ is the length of the actual argument list, written $|\es|=n$.
+When consider an argument list of length $n$, written $|\es|=n$, the opcon definition will receive a static list of length $n$ where the $j$th entry is the argument interface $(\slam{\kty}{\svar{ty}}{\sana{j}{\svar{ty}}}, \slam{\kunit}{\_}{\ssyn{j}})$.
+This \emph{argument interface list} is generated by the judgement $\keyw{args}(n)=\stx{args}$.
 
@@
 
-Recall that the kinding judgement is indexed by $n$.
-This is an upper bound on the argument index of terms of the form $\sana{n}{\st}$ and $\ssyn{n}$, enforced in Figure \ref{fig:kinding-ana-syn}.
+The index $n$ on the kinding judgement is an upper bound on the argument index of terms of the form $\sana{n}{\st}$ and $\ssyn{n}$, enforced in Figure \ref{fig:kinding-ana-syn}.
 Thus, if $\keyw{args}(n)= \stx{args}$ then $\sofkn{\emptyset}{\emptyset}{\Phi}{n}{\stx{args}}{\kargs}$.
-The rule (ocstruct-intro) ruled out writing either of these forms explicitly in an opcon definition by checking against the bound $n=0$.
-This is to prevent out-of-bounds errors: tycon providers can only access these forms via the argument interface list, which must have the correct length.
-
-\begin{figure}\vspace{-5px}
+The rule (ocstruct-intro) ruled out writing either of these forms explicitly in an opcon definition by checking against bound $n=0$.
+This is to prevent out-of-bounds errors: tycon providers can only access these forms via the argument interface list, which has the correct length.
+
+\begin{figure}\vspace{-10px}
 \begin{mathpar}\small
@@
     \sofkX{\ssyn{n'}}{\kty\times\kitm}
-}\vspace{-6px}
+}\vspace{-10px}
 \end{mathpar}
@@
 
-For $\sana{n}{\st}$, after normalizing $\st$, the argument environment, which contains the arguments themselves and the typing and tycon contexts, $\argEnv ::= \es; \Upsilon; \Phi$, is consulted to retrieve the $n$th argument and analyze it against $\st$.
+The static dynamics of $\sana{n}{\st}$ are interesting.
+After normalizing $\st$, the argument environment, which contains the arguments themselves and the typing and tycon contexts, $\argEnv ::= \es; \Upsilon; \Phi$, is consulted to analyze the $n$th argument against $\st$.
 If this succeeds, $\sqitm{\anatrans{n}{\st}}$ is generated:
 \begin{mathpar}\small
-\inferrule[s-ana-success]{
+\inferrule[n-ana-success]{
     \sval{\st}{\es;\Upsilon;\Phi}\\
@@
 \begin{mathpar}\small
-\inferrule[s-ana-fail]{
+\inferrule[n-ana-fail]{
     \sval{\st}{\es;\Upsilon;\Phi}\\
@@
 
-The kinding rules also prevent these forms  from being well-kinded when $n = 0$. Like the form $\srep{\st}$, these  are retained in values of kind $\kitm$, as shown in $\qitm_\text{ex}$.
-
-\subsection{Translation Validation}
+The kinding rules also prevent these translational forms  from being well-kinded when $n = 0$ and, like $\srep{\st}$ in Sec.
+\ref{sec:type-translations}, they  are retained in values of kind $\kitm$.%, as shown in $\qitm_\text{ex}$.
+
+\subsection{Translation Validation}\label{sec:translation-validation}
 \noindent
-The judgement $\trvalidate{\argEnv}{c}{\qitm}{\st}{\itm}$, defined by a single rule in Figure \ref{fig:translation-validation} and appearing as the final premise of (ana-intro) and the other rules described below, is pronounced ``translational internal term $\qitm$ generated by an opcon associated with $c$ under argument environment $\argEnv$ for an operation with type $\st$ is valid, so translation $\iota$ is produced''.
-For example, $$\trvalidate{(title; \concstr{EXMPL 2015}); \Upsilon_\text{ex}; \Phi_\text{rstr}\Phi_\text{lprod}}{\tcvar{lprod}}{\qitm_\text{ex}}{{\stx{paper}}}{\iota_\text{ex}}$$
+The judgement $\trvalidate{\argEnv}{c}{\qitm}{\st}{\itm\moutput}$, defined by a single rule in Figure \ref{fig:translation-validation} and appearing as the final premise of (ana-intro) and the other rules described below, is pronounced ``translational internal term $\qitm$ generated by an opcon associated with tycon $c$ under argument environment $\argEnv$ for an operation with type $\st$ is valid, so translation $\iota$ is produced''.
+For example, $$\trvalidate{(title; \concstr{EXMPL 2015}); \Upsilon_\text{ex}; \Phi_\text{rstr}\Phi_\text{lprod}}{\tcvar{lprod}}{\qitm_\text{paper}}{{\stx{paper}}}{\iota_\text{paper}}$$
 
 \noindent
-The purpose of this rule is to check that the generated translation will be well-typed \emph{no matter what the translations of types other than those constructed by $c$ are}.
-This \emph{translation independence} property will be the key to our conservativity theorem in Sec.
+The purpose of translation validation is to check that the generated translation will be well-typed \emph{no matter what the translations of types other than those constructed by $c$ are}.
+%This \emph{translation independence} property will be the key to our conservativity theorem in Sec.
 \ref{metatheory}.
@@
 \ref{sec:selective-type-translation-abstraction}.
-In our running example, this is $\tau_\text{abs/paper}$, i.e.
+In our running example, this is $\tau_\text{paper/abs}$, i.e.
 $\alpha_0 \times (\alpha_1 \times \kunit)$.
@@
 
-The judgement $\edeabs{c}{\argEnv}{\qitm}{\memD}{\memG}{\itm\moutput}{\memD\moutput}{\memG\moutput}$, appearing as the next premise, relates a translational internal term $\qitm$ to an internal term $\itm$ called a \emph{selectively abstracted term translation}, because all references to the translation of an argument (having any type) are replaced with a corresponding variable and, as in Sec.
-\ref{sec:selective-type-translation-abstraction}, all references to the translation of a type constructed by an extension tycon other than the ``delegated tycon'' $c$ are replaced with a corresponding abstract type variable.
+The judgement $\edeabs{c}{\argEnv}{\qitm}{\memD}{\memG}{\itm\moutput}{\memD\moutput}{\memG\moutput}$, appearing as the next premise, relates a translational internal term $\qitm$ to an internal term $\itm$ called a \emph{selectively abstracted term translation}, because all references to the translation of an argument (having any type) are replaced with a corresponding variable, $x$, which  will be of the selectively abstracted type translation of the type of that argument.
+For our example, $\small\edeabs{\tcvar{lprod}}{(title; \texttt{"EXMPL 2015"});\Upsilon_\text{ex};\Phi_\text{rstr}\Phi_\text{lprod}}{\qitm_\text{paper}}{\memD_\text{paper/abs}}{\emptyset}{\iota_\text{paper/abs}}{\memD_\text{paper/abs}}{\memG_\text{paper/abs}}$ where $\itm_\text{paper/abs} := (x_0, (x_1, ()))$.
+
 The type translation store, $\memD$, discussed previously, and term translation store, $\memG$, track these correspondences.
 Term translation stores have  syntax $\memG ::= \emptyset ~|~ \memG, n : \st \leadsto \iota/x : \tau$.
-Each entry can be read ``argument $n$ having type $\st$ and translation $\iota$ appears as variable $x$ with type $\tau$'', e.g.
-$\small\edeabs{\tcvar{lprod}}{(title; \texttt{"EXMPL 2015"});\Upsilon;\Phi_\text{rstr}\Phi_\text{lprod}}{\qitm_\text{ex}}{\memD_\text{paper/abs}}{\emptyset}{\iota_\text{ex/abs}}{\memD_\text{paper/abs}}{\memG_\text{ex/abs}}$ where $\itm_\text{ex/abs} := (x_0, (x_1, ()))$ and $\memG_\text{ex/abs} := 0 : \stx{title} \leadsto title/x_0 : \alpha_0, 1 : \stx{conf} \leadsto \texttt{"EXMPL 2015"}_\text{IL}/x_1 : \alpha_1$ and $\memD_\text{paper/abs}$ is from Sec.
+Each entry can be read ``argument $n$ having type $\st$ and translation $\iota$ appears as variable $x$ with type $\tau$''.
+In the example above, $\memG_\text{paper/abs} := 0 : \stx{title} \leadsto title/x_0 : \alpha_0, 1 : \stx{conf} \leadsto \texttt{"EXMPL 2015"}_\text{IL}/x_1 : \alpha_1$.
+
+To derive this, the judgement proceeded recursively along shared forms, as in Sec.
 \ref{sec:selective-type-translation-abstraction}.
-
-This judgement proceeds recursively along shared forms, like the selectively abstracted type translation judgement in Sec.
-\ref{sec:selective-type-translation-abstraction}.
-The key rule for references to argument translations derived via analysis is below ($\syntrans{n}$ is analagous; the full rules are in the supplement).% Note that we are rederiving the translation already determined in (s-ana-success) for simplicity (in practice, this might be cached):
+The interesting rule for argument translations derived via analysis is below ($\syntrans{n}$ is analagous; see supplement):% Note that we are rederiving the translation already determined in (s-ana-success) for simplicity (in practice, this might be cached):
 \begin{mathpar}\small
@@
 
-The third premise of (validate-tr) generates the type substitution and type formation contexts from the final type translation store,  $\memD' \leadsto \delta_\text{paper/abs} : \Delta_\text{paper/abs}$ as described in Sec.
+The third premise of (validate-tr) generates the type substitution and type formation contexts implied by the final type translation store as described in Sec.
 \ref{sec:selective-type-translation-abstraction}.
-Like type translation stores, each term translation store induces an internal term substitution,  $\gamma ::= \emptyset ~|~ \gamma, \iota/x$, and corresponding internal typing context $\Gamma$ by the judgement $\memG \leadsto \gamma : \Gamma$, appearing as the fourth premise.
+Similarly, each term translation store $\memG$ implies an internal term substitution,  $\gamma ::= \emptyset ~|~ \gamma, \iota/x$, and corresponding $\Gamma$ by the judgement $\memG \leadsto \gamma : \Gamma$, appearing as the fourth premise.
 Here, $\small
-\gamma_\text{ex/abs} := title/x_0, \texttt{"EXMPL 2015"}_\text{IL}/x_1$ and $\small\Gamma_\text{ex/abs} := x_0{:}\alpha_0, x_1{:}\alpha_1$.
-
-Finally, the fifth premise checks the abstracted term translation against the abstracted type translation according to the internal statics.
-Here, $\Delta_\text{paper/abs}~\Gamma_\text{ex/abs} \vdash \iota_\text{ex/abs} : \tau_\text{ex/abs}$, i.e.:
-\[\small(\alpha_0, \alpha_1)~(x_0 : \alpha_0, x_1 : \alpha_1) \vdash (x_0, (x_1, ())) : \alpha_0 \times (\alpha_1 \times \kunit)\]
-The result is that the translation of the labeled product $e_\text{ex}$ generated by $\tcvar{lprod}$ is checked with the references to term and type translations of regular strings replaced by variables and type variables, respectively.
-But because our  definition treated arguments parametrically, the check succeeds.
-Had we attempted to smuggle in a regular string that violated the regular string invariant, e.g.
-generating $\qitm_\text{bad} := (\concstr{}, (\concstr{}, ())$, the check fails (even though $(\concstr{}, (\concstr{}, ()) : \keyw{str} \times \keyw{str} \times \kunit$).% We  describe  an ill-behaved operator in Sec.
+\gamma_\text{paper/abs} := title/x_0, \texttt{"EXMPL 2015"}_\text{IL}/x_1$ and $\small\Gamma_\text{paper/abs} := x_0:\alpha_0, x_1:\alpha_1$.
+
+The critical fifth premise checks the selectively abstracted term translation $\iota_\text{paper/abs}$ against the selectively abstracted type translation $\tau_\text{paper/abs}$ under these contexts via the internal statics.
+Here, $\Delta_\text{paper/abs}~\Gamma_\text{paper/abs} \vdash \iota_\text{paper/abs} : \tau_\text{paper/abs}$, i.e.:
+\[\small\begin{array}{l}(\alpha_0, \alpha_1)~(x_0 : \alpha_0, x_1 : \alpha_1) \vdash (x_0, (x_1, ())) : \alpha_0 \times (\alpha_1 \times \kunit)\end{array}\]
+In summary, the translation of the labeled product $e_\text{paper}$ generated by $\tcvar{lprod}$ is checked with the references to term and type translations of regular strings replaced by variables and type variables, respectively.
+But because the  definition treated arguments parametrically, the check succeeds.
+% We  describe  an ill-behaved operator in Sec.
 \ref{metatheory}.
 
-Applying the substitutions $\gamma_\text{ex/abs}$ and $\delta_\text{paper/abs}$ in the conclusion of the rule, we arrive at the actual term translation $\iota_\text{ex}$, defined previously.
-Note that $\iota_\text{ex}$ has type $\tau_\text{paper}$ under the translation of $\Upsilon_\text{ex}$, i.e.
-$\vdash \Upsilon_\text{ex} \leadsto \Gamma_\text{ex}$ where $\Gamma_\text{ex} := title : \keyw{str}$.
+Applying the substitutions $\gamma_\text{paper/abs}$ and $\delta_\text{paper/abs}$ in the conclusion of the rule, we arrive at the actual term translation $\iota_\text{paper} := (title, (\concstr{EXMPL 2015}_\text{IL}, ()))$.
+Note that $\iota_\text{paper}$ has type $\tau_\text{paper}$ under the translation of $\Upsilon_\text{ex}$, i.e.
+$\vdash \Upsilon_\text{ex}~\mathtt{ctx} \leadsto \Gamma_\text{ex}$ where $\Gamma_\text{ex} := title : \keyw{str}$.
+This relationship will always hold, and implies type safety (Sec.
+\ref{metatheory}).
+
+Had we attempted to ``smuggle out'' a value of regular string type that violated the regular string invariant, e.g.
+generating $\qitm_\text{bad} := (\concstr{}, (\concstr{}, ())$, it would fail, because even though $(\concstr{}_\text{IL}, (\concstr{}_\text{IL}, ()) : \keyw{str} \times \keyw{str} \times \kunit$, it is not the case that $(\concstr{}_\text{IL}, (\concstr{}_\text{IL}, ()) : \alpha_0 \times (\alpha_1 \times \kunit)$.
+We call this property \emph{translation independence}.
 
 \subsection{Generalized Targeted Operations} \label{sec:targops}
-All non-introductory operations go through another generalized form, in this case for \emph{targeted operations}, $\etarg{\opname{op}}{\sttmidx}{e_\text{targ}}{\es}$, where $\opname{op}$ ranges over opcon names, ${\sttmidx}$ is the term index, $e_\text{targ}$ is the \emph{target argument} and $\es$ are the remaining arguments.
+\noindent All non-introductory operations go through the form for \emph{targeted operations}, $\etarg{\opname{op}}{\sttmidx}{e_\text{targ}}{\es}$, where $\opname{op}$ is the opcon name, ${\sttmidx}$ is the term index, $e_\text{targ}$ is the \emph{target argument} and $\es$ are the remaining arguments.
 Concrete desugarings for this form include $e_\text{targ}.\opname{op}\langle\sttmidx\rangle(\es)$ (and variants where the term index or arguments are omitted), projection syntax for use by record-like types, $e_\text{targ}\opname{\#}\conclbl{lbl}$, which desugars to $\etarg{\opname{\#}}{\conclbl{lbl}}{e_\text{targ}}{\cdot}$, and $e_\text{targ} \cdot e_\text{arg}$, which desugars to $\etarg{\opname{conc}}{\striv}{e_\text{targ}}{e_\text{arg}}$.
-We show other desugarings, including case analysis, in the supplement.
+We show other desugarings, e.g.
+case analysis, in the supplement.
 
@@
 Figure \ref{fig:example-tycons} showed portions of the opcon signatures of $\tcvar{rstr}$ and $\tcvar{lprod}$.
-The opcons associated with $\tcvar{rstr}$ are taken directly from Fulton et al.'s specification of regular string types \cite{sanitation-psp14}, with the exception of $\opname{case}$, which generalizes case analysis as defined there to arbitrary string patterns, which we discuss in the supplement.
+The opcons associated with $\tcvar{rstr}$ are taken directly from Fulton et al.'s specification of regular string types \cite{sanitation-psp14}, with the exception of $\opname{case}$, which generalizes case analysis as defined there to arbitrary string patterns, based on the form of desugaring we show in the supplement.
 The opcons associated with $\tcvar{lprod}$ are also straightforward: $\opname{\#}$ projects out the row with the provided label and $\opname{conc}$ concatenates two labeled products (updating common rows with the value from the right argument).
-Note that both $\tcvar{rstr}$ and $\tcvar{lprod}$ can define concatenation without conflict.
+Note that both $\tcvar{rstr}$ and $\tcvar{lprod}$ can define concatenation.
 %Targeted operations use no new mechanisms, so we only show regular string concatenation; others are in the supplement.
@@
 Targeted opcon definitions additionally synthesize a type.
-The rule (ocstruct-targ) in Figure \ref{ocstruct} ensures that no tycon defines an opcon twice and that the opcon definitions are well-kinded.
-For example, the definition of $\tcvar{rstr}$'s  $\opname{conc}$ opcon is shown in Figure \ref{fig:example-conc}.
+The rule (ocstruct-targ) in Figure \ref{ocstruct} ensures that it is well-kinded.
+For example, the definition of $\tcvar{rstr}$'s  $\opname{conc}$ opcon is shown in Figure \ref{fig:example-conc} (others will be in the supplement).
 
@@
 We can ignore the type synthesized by the first because by definition it is a regular string type with type index $\svar{tyidx}$.
-We case analyze the second against $\tcvar{rstr}$, extracting its index regular expression if so and raising an error if not.
-We then synthesize the resulting regular string type, using the helper function $\svar{rxconcat} :: \krx \rightarrow \krx \rightarrow \krx$ which generates the synthesized type's  index by concatenating the indices of the argument's types, and finally the translation, using an internal helper function $sconcat : \keyw{str} \rightarrow \keyw{str} \rightarrow \keyw{str}$, the translational term for which we assume has been substituted in directly.
+We case analyze the second against $\tcvar{rstr}$, to extract its index regular expression (raising an error if it is not of regular string type).
+We then synthesize the resulting regular string type, using the helper function $\svar{rxconcat} :: \krx \rightarrow \krx \rightarrow \krx$ which generates the synthesized type's  index by concatenating the indices of the argument's types consistent with the specification in \cite{sanitation-psp14}.
+Finally the translation is generated using helper function $sconcat : \keyw{str} \rightarrow \keyw{str} \rightarrow \keyw{str}$, the translational term for which we assume has been substituted in directly.
 
 The last premise of (syn-targ) again performs translation validation as described above.
-The only difference relative to (ana-intro) is that that we check the term translation against the type translation of the synthesized type, but the delegated tycon is that of the type synthesized by the target argument.
+The only difference relative to (ana-intro) is that that we check the term translation against the synthesized type but the delegated tycon is that of the type synthesized by the target argument.
 
 \subsection{Operations Over Other Types}
-The rules (ana-intro-other) and (syn-targ-other) are used to introduce and simulate other operations on terms of a type constructed by an ``other'' tycon.
-In both cases, the term index, rather than the tycon context, directly specifies the static function that maps the arguments to a translation.
+\noindent
+The rules (ana-intro-other) and (syn-targ-other) are used to introduce and simulate targeted operations on terms of all types constructed by any ``other'' tycon.
+In both cases, the term index, rather than the tycon context, directly specifies the translational internal term to be used.
 %In all other respects, they are familiar.
@@
 \section{Metatheory}\label{metatheory}
-\noindent We will now state the key metatheoretic properties of @$\lambda$. The full proofs are in the supplement.
+\noindent We will now state the key metatheoretic properties of @$\lambda$. The proofs are in the supplement.
 
@@
 \begin{theorem}[Static Progress]\label{thm:static-progress}
-If $\sofkn{\emptyset}{\emptyset}{\Phi}{n}{\st}{\kappa}$ and $\vdash \Phi$ and $|\es|=n$ then $\sval{\st}{\es; \Upsilon; \Phi}$ or $\serr{\st}{\es; \Upsilon; \Phi}$ or $\sstep{\st}{\es; \Upsilon; \Phi}{\st'}$.
+If $\sofkn{\emptyset}{\emptyset}{\Phi}{n}{\st}{\kappa}$ and $|\es|=n$ then $\sstep{\st}{\es; \Upsilon; \Phi}{\st'}$ or $\sval{\st}{\es; \Upsilon; \Phi}$ or $\serr{\st}{\es; \Upsilon; \Phi}$.
 \end{theorem}
@@
 \begin{theorem}[Static Preservation]\label{thm:static-preservation}
-If $\sofkn{\emptyset}{\emptyset}{\Phi}{n}{\st}{\kappa}$ and $\vdash \Phi$ and $|\es|=n$ and $\vdash_\Phi \Upsilon \leadsto \Gamma$ and $\sstep{\st}{\es; \Upsilon; \Phi}{\st'}$ then $\sofkn{\emptyset}{\emptyset}{\Phi}{n}{\st'}{\kappa}$.
+If $\sofkn{\emptyset}{\emptyset}{\Phi}{n}{\st}{\kappa}$ and $\vdash \Phi$ and $\vdash_\Phi \Upsilon~\mathtt{ctx} \leadsto \Gamma$ and $\sstep{\st}{\es; \Upsilon; \Phi}{\st'}$ then $\sofkn{\emptyset}{\emptyset}{\Phi}{n}{\st'}{\kappa}$.
 \end{theorem}\noindent
@@
 \begin{theorem}[Type Synthesis]
-If $\vdash \Phi$ and $\vdash_\Phi \Upsilon \leadsto \Gamma$ and $\esynX{e}{\st}{\iota}$ then $\vdash_\Phi \st \leadsto \tau$ (and thus $\istype{\st}{\Phi}$). 
+If $\vdash \Phi$ and $\vdash_\Phi \Upsilon~\mathtt{ctx} \leadsto \Gamma$ and $\esynX{e}{\st}{\iota}$ then $\st~\mathtt{type}_\Phi$. 
 \end{theorem}
+
+%% Note in these that we need a context translation judgement $\ctxtransX{\Upsilon}{\Gamma}$ based on $\tytransX{\st}{\tau}$.
 \paragraph{Type Safety}
@@
 Type safety for the IL \cite{pfpl} then implies that evaluation cannot go wrong.
-To prove this, we must in fact prove a stronger theorem: that a term's translation has its type's translation (the analagous notion is \emph{type-preserving compilation} in type-directed compilers \cite{tarditi+:til-OLD}):% produced as the translation of the external type.
+To prove this, we must in fact prove a stronger theorem: that a term's translation has its type's translation under the typing context's translation (the analagous notion is \emph{type-preserving compilation} in type-directed compilers \cite{tarditi+:til-OLD}):% produced as the translation of the external type.
 Note that we need only to state our top-level theorems for the analysis judgement  because of the subsumption rule.
 
-\begin{theorem}[Type-Preserving Translation]
-If $\vdash \Phi$ and $\vdash_\Phi \Upsilon \leadsto \Gamma$ and $\vdash_\Phi \st \leadsto \tau$ and $\eanaX{e}{\st}{\iota}$ then $\emptyset \vdash \Gamma$ and $\emptyset \vdash \tau$ and $\emptyset~\Gamma \vdash \iota : \tau$.
+\vspace{-4px}\begin{theorem}[Type-Preserving Translation]
+\label{thm:type-preserving-translation}
+If $\vdash \Phi$ and $\vdash_\Phi \Upsilon~\mathtt{ctx} \leadsto \Gamma$ and $\vdash_\Phi \st~\mathtt{type} \leadsto \tau$ and $\eanaX{e}{\st}{\iota}$ then $\emptyset \vdash \Gamma$ and $\emptyset \vdash \tau$ and $\emptyset~\Gamma \vdash \iota : \tau$.
 \end{theorem}
@@
 The interesting cases are (ana-intro), (ana-intro-other), (syn-trans) and (syn-trans-other); the latter two arise via subsumption.
-The result follows directly from the final premise of each rule, combined with lemmas that state that if $\memD \leadsto \delta : \Delta_\text{abs}$ and $\memG \leadsto \gamma : \Gamma_\text{abs}$ then $\emptyset \vdash \delta : \Delta_\text{abs}$ and $\Delta_\text{abs}~\Gamma \vdash \gamma : \Gamma_\text{abs}$, i.e.
-that all variables in $\Delta_\text{abs}$ and $\Gamma_\text{abs}$ have well-formed/well-typed substitutions in $\delta$ and $\gamma$,  and so applying them gives a well-typed term.
+The result follows directly from the translation validation process, combined with lemmas that state that all variables in $\Delta_\text{abs}$ and $\Gamma_\text{abs}$ in (tr-validate) have well-formed/well-typed substitutions in $\delta$ and $\gamma$,  so applying in the conclusion them gives a well-typed term.
 \end{proof-sketch}
@@
 
-\paragraph{Stability}\todo{if space, bring unicity back}
-Extending the tycon context does not change the meaning of any terms that were previously well-typed.
-\begin{theorem}[Stability]
-Letting $\Phi' := \Phi, \tcdef{\tc}{\psi}{\theta}$, if $\vdash \Phi'$ and $\vdash_\Phi \Upsilon \leadsto \Gamma$ and $\vdash_\Phi \st \leadsto \tau$ and $\eanaX{e}{\st}{\iota}$ then $\vdash_{\Phi'} \Upsilon \leadsto \Gamma$ and $\vdash_{\Phi'} \st \leadsto \tau$ and $\eana{\Upsilon}{\Phi'}{e}{\st}{\iota}$.
-\end{theorem}
+% \paragraph{Stability}\todo{if space, bring unicity back}
+% Extending the tycon context does not change the meaning of any terms that were previously well-typed.
+% \begin{theorem}[Stability]
+% Letting $\Phi' := \Phi, \tcdef{\tc}{\psi}{\theta}$, if $\vdash \Phi'$ and $\vdash_\Phi \Upsilon \leadsto \Gamma$ and $\vdash_\Phi \st \leadsto \tau$ and $\eanaX{e}{\st}{\iota}$ then $\vdash_{\Phi'} \Upsilon \leadsto \Gamma$ and $\vdash_{\Phi'} \st \leadsto \tau$ and $\eana{\Upsilon}{\Phi'}{e}{\st}{\iota}$.
+% \end{theorem}
 
 \paragraph{Conservativity} 
-Extending the tycon context also conserves all \emph{tycon invariants} maintained in any smaller tycon context. An example of a tycon invariant is the following:
+Extending the tycon context also conserves all \emph{tycon invariants} maintained in the original tycon context. An example of a tycon invariant is the following:
 
@@
 \end{tyconinvariant}
-\begin{proof-sketch} The proof is not unusually difficult because we have fixed the tycon context $\Phi_\text{rstr}$, so we can essentially treat the calculus like a type-directed compiler for a calculus with only two tycons, $\rightharpoonup$ and $\tcvar{rstr}$, plus one ``other'' one.
+\begin{proof-sketch} We have fixed the tycon context $\Phi_\text{rstr}$, so we can essentially treat the calculus like a type-directed compiler for a calculus with only two tycons, $\rightharpoonup$ and $\tcvar{rstr}$, plus some ``other'' one.
 Such a calculus and compiler specification was given in \cite{sanitation-psp14}, so we must simply show that the opcon definitions in $\tcvar{rstr}$ adequately satisfy these specification using standard techniques for the SL, a simply-typed functional language \cite{conf/pldi/Chlipala07}.
-The only ``twist'' is that the rule (syn-targ-other) can synthesize a regular string type.
-But if so, $\iota_\text{abs}$ will be checked against $\tau_\text{abs}=\alpha$.
-Thus, the invariants cannot be violated by direct application of relational parametricity in the IL (i.e.
-a ``free theorem'') \cite{WadlerThms}.
+The only ``twist'' is that the rule (syn-targ-other) can synthesize a regular string type paired with any translational term $\qity$.
+But it will be validated against $\tau_\text{abs}=\alpha$ because rule (abs-ext-not-delegated-new) applies in that case.
+Thus, the invariants cannot be violated by direct application of parametricity in the IL (i.e.
+this case can always be dispatched via a ``free theorem'') \cite{WadlerThms}.
 \end{proof-sketch}
 
-The reason why (syn-targ-other) is never a problem in proving a tycon invariant -- \emph{translation independence} of tycons -- turns out to be the same reason extending the tycon context conserves all tycon invariants.
+Another way to interpret this argument is that ``other'' types simulate all types that might arise from ``the future'' in that they can have any valid type translation (given directly in the type index) and their operators can produce any valid term translation (given directly in the term index).
+Because they are distinct from all ``present'' tycons, our translation validation procedure  ensures that they can be reasoned about uniformly -- they cannot possibly violate present invariants because they do not even know what type of term they need to generate.
+The only way to generate a term of type $\alpha$ is via an argument, which inductively obeys all tycon invariants.
+
+% The reason why (syn-targ-other) is never a problem in proving a tycon invariant -- \emph{translation independence} of tycons -- turns out to be the same reason extending the tycon context conserves all tycon invariants.
 A newly introduced tycon defining a targeted operator that synthesizes a regular string type, e.g.
@@
 
- We can state this more generally:
-
-\begin{theorem}[Conservativity] If $\vdash \Phi$ and $\tc \in \text{dom}(\Phi)$ and a tycon invariant for $\tc$ holds under $\Phi$: \begin{itemize}
-\item For all $\Upsilon, e, \sttyidx$, if $\eana{\Upsilon}{\Phi}{e}{\sty{\tc}{\sttyidx}}{\iota}$ and $\vdash_\Phi \Upsilon \leadsto \Gamma$ and $\vdash_\Phi {\sty{\tc}{\sttyidx}} \leadsto \tau$  then $P(\Gamma, \sttyidx, \iota)$.
-\end{itemize} then for all $\Phi' = \Phi, \tcdef{\tc'}{\psi'}{\theta'}$ such that $\vdash \Phi'$, the same tycon invariant holds under $\Phi'$: \begin{itemize}
+\begin{theorem}[Conservativity]\label{thm:conservativity} If $\vdash \Phi$ and $\tc \in \text{dom}(\Phi)$ and a tycon invariant for $\tc$ holds under $\Phi$: \begin{itemize}
+\item For all $\Upsilon, e, \sttyidx$, if $\eana{\Upsilon}{\Phi}{e}{\sty{\tc}{\sttyidx}}{\iota}$ and $\ctxtransX{\Upsilon}{\Gamma}$ and $\vdash_\Phi {\sty{\tc}{\sttyidx}} \leadsto \tau$  then $P(\Gamma, \sttyidx, \iota)$.
+\end{itemize} then for all $\Phi' = \Phi, \tcdef{\tc'}{\tcsig{\kappa'}{\chi'}}{\theta'}$ such that $\vdash \Phi'$, the same tycon invariant holds under $\Phi'$: \begin{itemize}
 \item For all $\Upsilon, e, \sttyidx$, if $\eana{\Upsilon}{\Phi'}{e}{\sty{\tc}{\sttyidx}}{\iota}$ and $\vdash_{\Phi'} \Upsilon \leadsto \Gamma$ and $\vdash_{\Phi'}{\sty{\tc}{\sttyidx}}\leadsto{\tau}$ then $P(\Gamma, \sttyidx, \iota)$.
@@
 \begin{proof-sketch}
-The proof maps every well-typed term under $\Phi'$ to a well-typed term under $\Phi$ with the same translation, and if the term has a type constructed by a tycon in $\Phi$, e.g.
+Our proof of the more general property is a realization of this intuition that ``other'' types simulate future types.
+We simply map every well-typed term under $\Phi'$ to a well-typed term under $\Phi$ with the same translation, and if the term has a type constructed by a tycon in $\Phi$, e.g.
 $\tc$, the new term has a type constructed by that tycon with the same type translation, and only a slightly different type index.
-In particular, the mapping's effect on static terms is to replace all types constructed by $\tc'$ with a type constructed by $\keyw{other}[m;\ktyidx']$ for some $m$ corresponding to $\tc'$ and the index kind of $\tc'$.
-If $P(\Gamma, \st, \iota)$ is preserved under this transformation then we can simply invoke the existing proof of the tycon invariant.
+In particular, the mapping's effect on static terms is to replace all types constructed by $\tc'$ with a type constructed by $\keyw{other}[m;\ktyidx']$.
+If $P(\Gamma, \sttyidx, \iota)$ is preserved under this transformation on $\sttyidx$ then we can simply invoke the existing proof of the tycon invariant.
 We call such propositions \emph{modular}.
 Non-modular propositions are uninteresting because they distinguish tycons ``from the future''.
-
-On external terms, the mapping replaces all intro and targeted terms associated with $\tc'$ with an equivalent one that passes through the rules (ana-intro-other) and (syn-targ-other) by partially applying the intro and targeted opcon definitions to generate the term indices.
-Typing, kinding, static normalization and selective translation abstraction are preserved under the mapping, defined inductively in the supplement.
-Note that for this reason all propositions decidable by the SL are modular.
-\end{proof-sketch}
+Our regular string proposition is clearly modular because regular expressions don't contain types at all.
+
+On external terms, the mapping replaces all intro and targeted terms that delegated to $\tc'$ with equivalent ones that pass through rules (ana-intro-other) and (syn-targ-other) by pre-applying the intro and targeted opcon definitions to generate the term indices.\end{proof-sketch}
 %Had the translational term generated by the intro opcon definition been, e.g., $(\anatrans{0}{\stx{title}}, (\texttt{"TEST"}, ())$, then the corresponding abstract term would be $(x_0, (\texttt{"TEST"}, ())$ and the check would fail because $(\alpha_0, \alpha_1)~(x_0 : \alpha_0) \nvdash \texttt{"TEST"} : \alpha_1$.
@@
 \section{Related Work and Discussion}\label{prior-work}
+\noindent We are not the first to use a semantics distinguishing the EL from a smaller IL  for a language specification.
+For example, the Harper-Stone semantics for Standard ML takes a similar approach, though the EL and IL are governed by a common, and fixed, type system there \cite{Harper00atype-theoretic}.
+Our specification style is also comparable to that of the initial stage of a type-directed compiler, e.g.
+the TIL compiler for Standard ML \cite{tarditi+:til-OLD}, here lifted ``one level up'' into the semantics of the language itself and made extensible.
+%Our work shows ho style of semantics can  support extensions.
+
 %\paragraph{Term Rewriting}
-Language-integrated static term rewriting systems, like Template Haskell \cite{SheardPeytonJones:Haskell-02} and Scala's static macros \cite{ScalaMacros2013}, can be used to decrease complexity when an isomorphic embedding into the underlying type system is possible.
+Language-integrated static term rewriting systems, like Template Haskell \cite{SheardPeytonJones:Haskell-02} and Scala's static macros \cite{ScalaMacros2013}, can be used to decrease complexity when an isomorphic embedding into the underlying type system is already possible.
 % If each metaprogram is invoked like a function, inner macros are expanded before outer macros, and the mechanism enforces abstraction barriers by ensuring that the rewriting logic is hygienic, cannot modify surrounding code, and does not depend on shared state, then compositional reasoning is possible: the meaning of a term depends only on its subterms, not on the specific position it appears in a program.% Complex macros can, however, still be difficult to reason about, so many static macro systems impose further constraints (e.g.
@@
 %Rewriting systems that permit global pattern-based dispatch, however, do not admit strong modular reasoning principles (the same term might match multiple patterns, defined separately, creating ambiguities).
-Similarly, when an embedding that preserves a desired static semantics exists, but a different  embedding preserves the cost semantics, term rewriting  can also be used to perform ``optimizations'', achieving an isomorphism.
+Similarly, when an embedding that preserves a fragment's static semantics exists, but a different  embedding better approaches a desired cost semantics, term rewriting  can also be used to perform ``optimizations''.
 Care is needed when this changes the type of a term.
-Type abstraction has been used for  this purpose in \emph{lightweight modular staging} (LMS) \cite{Rompf:2012:LMS}.
-In both cases, the type system is fixed (e.g.
-in LMS, Scala's).
+Type abstraction has been used for this purpose in work on \emph{lightweight modular staging} (LMS) \cite{Rompf:2012:LMS}.
+In all of these cases, the type system remains fixed.
 
@@
 Were it to be, the system could be seen in ways as a degenerate mode of use of our work: we further cover the cases when new  operators are needed. 
-For example, labeled tuples couldn't be seen as refinements of nested pairs because the row projection operators don't exist.
+For example, labeled tuples couldn't be seen as refinements of nested pairs because label-indexed row projection operators simply don't exist.
 %Labeled products are not refinements of binary products because the label projection operator is not defined for binary products.%Regular string types could also support an operation that the internal string type would not (e.g.
@@
 
-Many \emph{language frameworks} exist that can simplify  dialect implementation (cf.
+Many \emph{language frameworks} exist that simplify  dialect implementation (cf.
 \cite{erdweg2013state}).
@@
 Fewer tools require knowledge of all external tycons in a typed translation semantics.
-As discussed previously, our treatment of concrete syntax in both the EL and SL defers to recent work on \emph{type-specific languages}, which takes a similar bidirectional approach for composably introducing syntactic desugarings \cite{TSLs}.
 Some language frameworks do address the EP, e.g.
 by encoding terms and types as open datatypes \cite{conf/ppdp/LohH06}, but this makes it quite difficult to reason modularly, particularly about metatheoretic properties specific to typed languages, like type safety and tycon invariants.
-Our key insight is to instead associate term-level opcons with tycons, which then become the fundamental constituents of the semantics (consistent with Harper's informal notation \cite{pfpl}).
+Our key insight is to instead associate term-level opcons with tycons, which then become the fundamental constituents of the semantics (consistent with Harper's informal notation from Sec.
+\ref{intro}).
+
+As discussed, our treatment of concrete syntax defers to recent work on \emph{type-specific languages}, which takes a similar split bidirectional approach for composably introducing syntax \cite{TSLs}.
+We focus on  semantics.
 
@@
 canonical forms) \cite{conf/popl/DelawareOS13}.
-Several authors, e.g. Chlipala  \cite{Chlipala10}, suggest  proof automation as a heuristic solution to the problem.
-
-In contrast, in @$\lambda$, fragment providers need not provide the semantics with mechanized proofs to benefit from rigorous modular reasoning principles.
+Several authors, notably Chlipala  \cite{Chlipala10}, suggest  proof automation as a heuristic solution to the problem.
+
+In contrast, in @$\lambda$, fragment providers need not provide the semantics with mechanized specfications or proofs to benefit from rigorous modular reasoning principles.
 Instead, under a fixed tycon context, the calculus can be reasoned about like a very small type-directed compiler \cite{tarditi+:til-OLD,conf/pldi/Chlipala07}.
-Errors in reasoning can only lead to failure at typechecking time, via a novel form of \emph{translation validation} \cite{Pnueli-Siegel-Singerman98}.
-Incorrect opcon definitions (relative to a specification, e.g.
+Errors in reasoning can only lead to failure at typechecking time, via our chief contribution: a novel form of \emph{translation validation} \cite{Pnueli-Siegel-Singerman98}.
+Incorrect opcon definitions relative to a specification (e.g.
 \cite{sanitation-psp14} for regular strings) can at worst weaken expected invariants at that tycon, like incorrectly implemented modules in ML.
-Thus, modular tycons can reasonably be tested ``in the field'' without concern about the reliability of the system as a whole.
-To eliminate even these localized failure modes for ``reliability-critical'' tycons, we plan to introduce \emph{optional} proof mechanization into the SL (by basing it on a dependently typed language like Coq).
+Thus, modular tycons can reasonably be tested ``in the field'' without concern about the reliability of the semantics as a whole.
+To eliminate even these localized failure modes, we plan to introduce \emph{optional} specification and proof mechanization into the SL (by basing it on a dependently typed language like Coq, rather than ML).
+Because types are values in the SL, the SL can be seen as building on the concept of \emph{type-level computation}.
 %We also hope to mechanize the metatheory of @$\lambda$ itself in future work.%Every combination of fragments is a new dialect, and must be reasoned about monolithically.
@@
 Indeed, proofs of tycon invariants can rely on existing parametricity theorems \cite{WadlerThms}.
-Our work is reminiscent of  work on elaborating an ML-style module system into System $\mathbf{F}_\omega$ \cite{conf/tldi/RossbergRD10}.
+Our work is thus reminiscent of  work on elaborating an ML-style module system into System $\mathbf{F}_\omega$ \cite{conf/tldi/RossbergRD10}.
 Unlike in module systems, type translations (analagous to the choice of representation for an abstract type) are statically \emph{computed} based on a type index, rather than statically \emph{declared}.
@@
 In contrast, modules and ADTs can only specify a fixed number of operations, and each must have function type.
-Note however that these are not competing mechanisms: we did not specify quantification over external types here for simplicity, but we conjecture that it is complementary and thus @$\lambda$ could serve as the core of a language with an ML-style module system.
-Another related direction is \emph{tycon functors}, which would abstract over tycons with the same signature to support tunable cost semantics.
+Note also that these are not competing mechanisms: we did not specify quantification over external types here for simplicity, but we conjecture that this is entirely complementary and thus @$\lambda$ could serve as the core of a language with an ML-style module system and parametric polymorphism.
+Another related direction is to explore \emph{tycon functors}, which would abstract over tycons with the same signature to support modularly tunable cost semantics.
 % This permits us tocode targeting a separate IL, and permit operations whose types cannot simply be written as functions.
@@
 
+\section{Conclusion}
+\noindent We have specified a remarkably simple but surprisingly powerful typed translation semantics, @$\lambda$, where new type constructors can be introduced ``from within''.
+The corresponding operator constructors determine types and translations ``actively'', i.e.
+using static functions.
+A simple form of translation validation based on existing notions of type abstraction  ensures that opcons associated with one tycon maintain translation independence from all others, guaranteeing that a wide class of important properties can be reasoned about modularly.
+Implementation is ongoing.
+
 A limitation of our approach is that it supports only  fragments with the standard ``shape'' of typing judgement.
 Fragments that require new forms of scoped contexts (e.g.
-symbol contexts \cite{pfpl}) or unscoped declarations cannot presently be defined.
-Relatedly, the language controls variable binding, so, for example, linear type systems, cannot be defined.
+symbol contexts \cite{pfpl}) cannot presently be defined.
+Relatedly, the language controls variable binding, so, for example, linear  type systems cannot be defined.
 Another limitation is that opcons cannot directly invoke one another (e.g.
 a \opname{len} opcon on regular strings could not construct a natural number).
-We conjecture that these are not fundamental limitations and expect @$\lambda$ to serve as a minimal foundation for future efforts that increase  expressiveness while maintaining the strong guarantees, like type safety and conservativity, established  here.
+We conjecture that these are not fundamental limitations and expect @$\lambda$ to serve as a  foundation for future efforts that increase  expressiveness while maintaining the strong modularity guarantees established  here.
 %(tycons could define new contexts which are threaded opaquely through ``outside'' operations).
@@
 
-\bibliographystyle{abbrv}
+\bibliographystyle{abbrvnat}
 

### 2014-11-17 sanitation-psp14/tr/strings_chapter.tex
Looks like a translation correctness proof
and some other minor changes.

@@
 \newcommand{\trden}[1]{\llbracket #1 \rrbracket} % = Translation Denotation.
-
 % Relations
@@
 \title{Statically Typed String Sanitation Inside a Python: Technical Report}
-\author{Nathan Fulton \\ Cyrus Omar \\ Jonathan Aldrich}
+\author{Nathan Fulton\and Cyrus Omar\and Jonathan Aldrich}
 
@@
 
+Theorems and lemmas appearing in \cite{fulton2014} are numbered, while
+supporting facts appearing only in the Technical Report are lettered.
+Numbered items correspond to the numbering in \cite{fulton2014}.
+
 \section{Proofs of Lemmas and Theorems about $\lambdas$}
@@
 this section are proven as stated in \cite{fulton2014}.
-
-Theorems and lemmas appearing in \cite{fulton2014} are numbered, while
-supporting facts appearing only in the Technical Report are lettered.
 
@@
 
+\begin{thm}[Safety for $\lambdap$] If $\emptyset \vdash \iota : \tau$ 
+  then $\iota \sreduces \dot v$ and $\emptyset \vdash \dot v : \tau$.
+\end{thm}
+
+We can also define canonical forms for regular expressions and strings in the usual way:
+
+\begin{lem}[Canonical Forms for Target Language]
+~
+\begin{itemize}
+\item 
+If $\emptyset \vdash \iota : \regex$ then $\iota \treduces \rx{r}$ such that $r$ is a well-formed regular expression. 
+\item 
+If $\emptyset \vdash \iota : \str$ then $\iota \treduces \tstr{s}$.
+\end{itemize}
+\end{lem}
+
 \section{Proofs and Lemmas and Theorems About Translation}
+
+\begin{thm}[Translation Correctness]\label{thm:trcorrect}
+  If $\tctx \vdash e : \sigma$ then 
+  there exists an $\iota$ such that $\trden{e} = \iota$
+  and $\trden{\tctx} \vdash \iota : \trden{\sigma}$.
+  Furthermore, $e \sreduces v$ and
+  $\iota \treduces \dot{v}$ such that
+  $\trden{v} = \dot{v}$.
+\end{thm}
+\begin{proof}
+We present a proof by induction on the derivation that $\tctx \vdash e : \sigma$.
+we write $e \leadsto \iota$ as shorthand for the final property.
+
+\begin{itemize}[label=$ $, itemsep=1ex]
+\item \textbf{Case $e = \rstr{s}$}. Suppose $\tctx \vdash \rstr{s} : \sigma$.
+
+By examination the syntactic structure of conclusions in the relation S-T, 
+we know this is true just in case $\sigma = \stringin{r}$ for some $r$ such that
+$s \in \lang{r}$; and of course, there is always such an $r$.
+
+There are no free variables in $\rstr{s}$, so we might as well proceed from the
+fact that $\emptyset \vdash \rstr{s} : \stringin{r}$.
+
+By definition of the translation ($\trden{\cdot}$) the following statements hold:
+\begin{align}
+\trden{\rstr{s}} &= \str{s} \\
+\trden{\stringin{r}} &= \str \\
+\trden{\emptyset} &= \emptyset
+\end{align}
+Note that $\emptyset \vdash \str{s} : \str$ by P-T-Str.
+Recall that contexts are standard and, in particular, can be weakened.
+So since $\trden{\tctx}$ is either a weakening of $\emptyset$ or $\emptyset$ itself,
+$\trden{\tctx} \vdash \tstr{s} : \str$ by weakening.
+
+Summarily, $\str{s}$ is a term of $\lambdap$ such that $\trden{\tctx} \vdash \str{s} : \trden{\sigma}$
+
+It remains to be shown that there exist $v, \dot{v}$ such that $\rstr{s} \sreduces v$,
+$\str{s} \treduces \dot{v}$, and $\trden{v} = \dot{v}$. But this is immediate because
+each term evaluates to itself and we have already established the equality.
+
+\item \textbf{Case $e = \rconcat{e_1}{e_2}$}. This case is an obvious appeal to induction.
+
+\item \textbf{Case $e = \strcase{e_1}{e_2}{x,y.e_3}$}. This case relies on our definition
+of context translation.
+
+Suppose $\sctx \vdash \strcase{e_1}{e_2}{x,y.e_3} : \sigma$.
+By inversion of the typing relation it follows that
+$\sctx \vdash e_1 : \stringin{r}$, $\sctx \vdash e_2 : \sigma$ and $\sctx, x : \stringin{\lhead{r}}, y : \stringin{\ltail{r}} \vdash e_3 : \sigma$.
+
+By induction, there exists an $\iota_1$ such that $\trden{e_1} = \iota_1$,
+$\trden{\sctx} \vdash \iota_1 : \trden{\sigma}$, and $e_1 \leadsto \iota_1$.
+Similarly for $e_2$ and some $\iota_2$.
+
+By canonical forms, $e_1 \sreduces \rstr{s}$ and so $\iota_1 \treduces \tstr{s}$ by $\leadsto$.
+
+Choose $\iota = \tconcat{\iota_1}{\iota_2}{x,y.\iota_3}$ \todo{and note that by the properties established via induction,
+$\trden{e} = \iota$ and $\trden{\sctx} \vdash \iota : \trden{\sigma}$.}
+
+Suppose $s = \epsilon$. Then $e \sreduces v$ where $e_2 \sreduces v$
+and $\iota \treduces \dot{v}$ where $\iota_2 \treduces \dot{v}$.
+But recall that $e_2 \leadsto v_2$ and so $\trden{v} = \dot{v}$.
+
+Suppose otherwise that $s = at$ for some character $a$ and string $t$.
+Then $e \sreduces v$ where $[a, t / x, y] e_3 \sreduces v$.
+Similarly, $\iota \treduces \dot{v}$ where $[a, t / x, y] \iota_3 \treduces \dot{v}$
+
+\end{itemize}
+\end{proof}
+
+\begin{thm}[Correctness of Input Sanitation for Translated Terms]\label{thm:main}
+  If $\trden{e} = \iota$ and $\emptyset \vdash e : \stringin{r}$ then $\iota \sreduces \tstr{s}$
+  for $s \in \lang{r}$.
+\end{thm}
+\begin{proof}
+  By Theorem \ref{thm:trcorrect} and the rules given, $\iota \sreduces \tstr{s}$ implies that $e \sreduces \strin{s}$.
+  Theorem \ref{thm:scorrect} together with the assumption that $e$ is well-typed implies that $s \in \lang{r}$.
+\end{proof}
 

### 2014-11-18 sanitation-psp14/tr/strings_chapter.tex
prep for camera ready.

@@
 \begin{abstract}
-This Technical Report contains supporting materials for Statically Typed
-String Sanitation Inside a Python \cite{fulton2014} including proofs of lemmas
-and theorems asserted in the paper, examples, and additional discussion of
-the paper's technical content.
+Web applications must ultimately command systems like web browsers and database engines using strings.
+Strings derived from improperly sanitized user input  can thus be a vector for command injection attacks.
+
+In this report, we introduce \emph{regular string types}, which classify strings known statically to be in a specified regular language.
+These types come equipped with common operations like concatenation, substitution and coercion, so they  can be used to implement, in a conventional manner, the portions of a web application or application framework that must directly construct command strings.
+Simple type annotations at key interfaces can be used to statically verify that sanitization has been performed correctly without introducing redundant run-time checks.
+We specify this type system in a minimal typed lambda calculus, $\lambda_{RS}$.
+
+We then specify a translation from $\lambda_{RS}$ to a language fragment containing only standard strings and regular expressions and prove that the correctness theorem for
+$\lambda_{RS}$ is preserved under translation.
 \end{abstract}
 
-Theorems and lemmas appearing in \cite{fulton2014} are numbered, while
-supporting facts appearing only in the Technical Report are lettered.
+\section{Introduction}
+
+Command injection vulnerabilities are among the most common and severe security vulnerabilities in modern web applications.
+They arise because web applications, at their boundaries, control external systems using commands represented as  strings.
+In \cite{fulton2014} the authors argue that extensible type systems are an attractive solution to this important security problem.
+This Technical Report contains supporting evidence for claims put forth and explained in \cite{fulton2014}, including proofs of lemmas
+and theorems asserted in the paper, examples, and additional discussion of the paper's technical content.
+
+Theorems and lemmas appearing in \cite{fulton2014} are numbered, while supporting facts appearing only in the Technical Report are lettered.
 Numbered items correspond to the numbering in \cite{fulton2014}.
@@
 The relation $\lhead{r}=r'$ is defined in terms of the structure of $r$:
-\begin{itemize}[noitemsep]
-\item $\lhead{\epsilon} = \epsilon$
-\item $\lhead{.} = a_1 + a_2 + ... + a_n$ for all $a_i \in \Sigma$ where $|\Sigma| = n$.
-\item $\lhead{a} = a$ %where $a \in \Sigma$
-\item $\lhead{r_1\cdot r_2} = \lhead{r_1}$
-\item $\lhead{r_1 +  r_2}  = \lhead{r_1} + \lhead{r_2}$
-\item $\lhead{r*} = \epsilon + \lhead{r}$
-\end{itemize}
+\begin{align*}
+\lhead{r}                 &= \lhead{r, \epsilon} \\
+\lhead{\epsilon, r'}      &= \epsilon \\
+\lhead{a, r'}             &= a \\
+\lhead{r_1\cdot r_2, r'}  &= \lhead{r_1, r_2} \\
+\lhead{r_1 + r_2, r'}     &= \lhead{r_1, r'} + \lhead{r_2, r'} \\
+\lhead{r^*, r'}           &= \lhead{r', \epsilon} + \lhead{r, \epsilon}
+\end{align*}
 \end{defn}
@@
 Note that $\lhead{r} = a_1 + a_2 + ... + a_i$.
-We define $\ltail{r} = \delta_{a_1}(r) + \delta_{a_2}(r) + ... + \delta_{a_i}(r)$.
+We define $\ltail{r} = \delta_{a_1}(r) + \delta_{a_2}(r) + ... + \delta_{a_i}(r) + \epsilon$.
 \end{defn}
+
+Using these definitions of head and tail, we establish a correctness
+result upon which type soundness for the concatenation operator in
+$\lambdas$ depends.
+
+\begin{trlem}[Leading characters are in the head] \label{trlemmax}
+If $c_1 \cdot c_2 \cdot ... \cdot c_m \in \lang{r}$, then
+$c_1 \in \lhead{r}$.
+\end{trlem}
+\begin{proof}
+By structural induction on $r$...
+\end{proof}
+
+\begin{trthm}[Correctness of \textsf{ltail}]
+If $s \in \lang{r}$ then $s \in \lang{\lhead{r}} \cdot \lang{\ltail{r}}$.
+\end{trthm}
+\begin{proof}
+The proof proceeds by structural induction on $r$.
+In each case, we demonstrate three facts:
+\begin{itemize}
+  \item First, that the string $s = a \cdot b$ is a concatenation of two substrings (each of which may be $\epsilon$).
+  \item Second, $a \in \lang{\lhead{r}}$ (or, equivalently, $s \in \lang{\lhead{r}}$ letting $b = \epsilon$).
+  \item Third, there exists a $c \in \lhead{r}$ such that $a = c \cdot a'$ and $a' \cdot b \in \lang{\delta_c(r)}$, or else $s = a$. 
+\end{itemize}
+
+Throughout, we consider $\epsilon$ as idenetity under all operators for expressions and $\cdot$ for strings.
+
+\begin{itemize}[label=$ $,itemsep=1ex]
+
+\item \textbf{Case $r = c$ for $c \in Sigma$}.
+Note that if $s \in \lang{c}$ then $s = c$.
+Note that $\lhead{c} = \lhead{c, \epsilon} = c$, and $c \in \lang{c}$.
+So $c \in \lang{c}$, which is sufficient since $\epsilon \in \ltail{r}$.
+
+\item \textbf{Case $r = r_1 \cdot r_2$}.
+Suppose $s \in \lang{r}$.
+We may decompose $s$ as $s_1 \cdot ... \cdot s_n$ such that $s_1 = c_1 \cdot ... \cdot c_m \in r_1$.
+Note that $c_1 \in \lhead{r_1}$ by \ref{trlemmax}.
+Let $a' = c_2 \cdot ... \cdot c_m$.
+It suffices to show that $a' \cdot b \in \lang{\delta_c(r)}$.
+Since $s = c_1 \cdot a' \cdot b$ and $s \in \lang{r}$, it follows by the definition of derivative that $a' \cdot b \in \delta_{c_1}(r)$.
+
+\item \textbf{Case $r = r_1 + r_2$}.
+Suppose $s \in \lang{r_1 + r_2}$
+so that $s \in \lang{r_1}$ or $s \in \lang{r_2}$.
+In either case, the result follows by induction.
+
+\item \textbf{Case $r = q^*$}.
+Note that $s = \epsilon$ or else $s = s_1 \cdot s_2 \cdot ... s_n$ where each $s_i \in \lang{q}$\footnote{intuitively, $s$ is either empty or a finitary concatenation of strings matching $q$.}.
+
+Suppose $s = \epsilon$.
+Note that $\lhead{r} = \lhead{q^*} = \lhead{q^*, \epsilon} = \lhead{\epsilon, \epsilon} + h' = \epsilon + h'$.
+Therefore, $s \in \lhead{r}$.
+
+In the other case, suppose $s = s_1 \cdot ... \cdot s_n$ where each $s_i \in \lang{q}$.
+Let $a = s_1$ and $b = s_2 \cdot ... s_n$, so that $s = a \cdot b$.
+Note that $a = c_1 \cdot ... \cdot c_n$.
+We will show that $c_1 \in \lang{\lhead{q^*}}$ and that $(c_2 \cdot ... c_n) \cdot b \in \delta_{c_1}(q^*)$.
+
+The latter property is trivial. Note that $s = c_1 \cdot c_2 \cdot ... \cdot c_n \cdot b \in \lang{q^*}$.
+The result follows immediately from the definition of derivative.
+
+What remains to be shown is that $c_1 \in \lang{q^*}$.
+Note that $\lhead{q^*} = \epsilon + \lhead{q, \epsilon}$.
+Recall that $a = s_1$ where $s_1 \in \lang{q}$.
+Also recall that $a= c_1 \cdot ... \cdot c_n$.
+Therefore, $c_1 \in \lhead{q}$ by \ref{trlemmax}.
+
+\end{itemize}
+\end{proof}
 
@@
 \begin{proof}
-\todo{Follows directly from progress and preservation.}
+Follows directly from progress and preservation.
 \end{proof}

### 2014-11-21 sanitation-psp14/tr/strings_chapter.tex
PSP14TR - fixed formating

@@
-\documentclass[12pt]{article}
-
-\usepackage{lineno}
-
-\usepackage{afterpage}
-\usepackage{listings}
-\usepackage[usenames,dvipsnames]{color}
-\usepackage{mathpartir}
+\documentclass[11pt,leqno]{article}
+\usepackage{cmu-titlepage2}
+\usepackage{fullpage}
+\usepackage[small,bf]{caption2}
+\usepackage{times}
+\usepackage{helvet}
+\usepackage{courier}
+
 \usepackage{amsmath}
 \usepackage{amssymb} 
-
-% Hack. Something's wrong with PLAS paper when using the ACM Proc docclass
-\let\proof\relax
-\let\endproof\relax
 \usepackage{amsthm}
-
-\usepackage[nomarkers,figuresonly]{endfloat}
-\renewcommand{\efloatseparator}{\vfill}
-
 \usepackage{ stmaryrd }
-\usepackage{verbatimbox}
-\input{../../att-icfp14/macros-atlam}
-\usepackage{alltt}
-\renewcommand{\ttdefault}{txtt}
-\usepackage{color}
+\usepackage{mathpartir}
+\usepackage{enumerate}
 \usepackage[usenames,dvipsnames,svgnames,table]{xcolor}
-\definecolor{mauve}{rgb}{0.58,0,0.82}
-\definecolor{light-gray}{gray}{0.5}
-\usepackage{listings}
-\usepackage{wasysym}
-    \makeatletter
-\usepackage{enumitem}
-\usepackage{enumerate}
-
-% \btIfInRange{number}{range list}{TRUE}{FALSE}
-%
-% Test if int number <number> is element of a (comma separated) list of ranges
-% (such as: {1,3-5,7,10-12,14}) and processes <TRUE> or <FALSE> respectively
-%
-        \newcount\bt@rangea
-        \newcount\bt@rangeb
-
-        \newcommand\btIfInRange[2]{%
-            \global\let\bt@inrange\@secondoftwo%
-            \edef\bt@rangelist{#2}%
-            \foreach \range in \bt@rangelist {%
-                \afterassignment\bt@getrangeb%
-                \bt@rangea=0\range\relax%
-                \pgfmathtruncatemacro\result{ ( #1 >= \bt@rangea) && (#1 <= \bt@rangeb) }%
-                \ifnum\result=1\relax%
-                    \breakforeach%
-                    \global\let\bt@inrange\@firstoftwo%
-                \fi%
-            }%
-            \bt@inrange%
-        }
-
-        \newcommand\bt@getrangeb{%
-            \@ifnextchar\relax%
-            {\bt@rangeb=\bt@rangea}%
-            {\@getrangeb}%
-        }
-
-        \def\@getrangeb-#1\relax{%
-            \ifx\relax#1\relax%
-                \bt@rangeb=100000%   \maxdimen is too large for pgfmath
-            \else%
-                \bt@rangeb=#1\relax%
-            \fi%
-        }
-
-%
-% \btLstHL{range list}
-%
-        \newcommand{\btLstHL}[1]{%
-            \btIfInRange{\value{lstnumber}}{#1}%
-            {\color{black!10}}%
-            {\def\lst@linebgrd}%
-        }%
-
-%
-% \btInputEmph[listing options]{range list}{file name}
-%
-        \newcommand{\btLstInputEmph}[3][\empty]{%
-            \lstset{%
-                linebackgroundcolor=\btLstHL{#2}%
-                \lstinputlisting{#3}%
-            }% \only
-        }
-
-% Patch line number key to call line background macro
-        \lst@Key{numbers}{none}{%
-            \def\lst@PlaceNumber{\lst@linebgrd}%
-            \lstKV@SwitchCases{#1}{%
-                none&\\%
-                left&\def\lst@PlaceNumber{\llap{\normalfont
-                \lst@numberstyle{\thelstnumber}\kern\lst@numbersep}\lst@linebgrd}\\%
-                right&\def\lst@PlaceNumber{\rlap{\normalfont
-                \kern\linewidth \kern\lst@numbersep
-                \lst@numberstyle{\thelstnumber}}\lst@linebgrd}%
-            }{%
-                \PackageError{Listings}{Numbers #1 unknown}\@ehc%
-            }%
-        }
-
-% New keys
-        \lst@Key{linebackgroundcolor}{}{%
-            \def\lst@linebgrdcolor{#1}%
-        }
-        \lst@Key{linebackgroundsep}{0pt}{%
-            \def\lst@linebgrdsep{#1}%
-        }
-        \lst@Key{linebackgroundwidth}{\linewidth}{%
-            \def\lst@linebgrdwidth{#1}%
-        }
-        \lst@Key{linebackgroundheight}{\ht\strutbox}{%
-            \def\lst@linebgrdheight{#1}%
-        }
-        \lst@Key{linebackgrounddepth}{\dp\strutbox}{%
-            \def\lst@linebgrddepth{#1}%
-        }
-        \lst@Key{linebackgroundcmd}{\color@block}{%
-            \def\lst@linebgrdcmd{#1}%
-        }
-
-% Line Background macro
-        \newcommand{\lst@linebgrd}{%
-            \ifx\lst@linebgrdcolor\empty\else
-                \rlap{%
-                    \lst@basicstyle
-                    \color{-.}% By default use the opposite (`-`) of the current color (`.`) as background
-                    \lst@linebgrdcolor{%
-                        \kern-\dimexpr\lst@linebgrdsep\relax%
-                        \lst@linebgrdcmd{\lst@linebgrdwidth}{\lst@linebgrdheight}{\lst@linebgrddepth}%
-                    }%
-                }%
-            \fi
-        }
-
- % Heather-added packages for the fancy table
- \usepackage{longtable}
- \usepackage{booktabs}
- \usepackage{pdflscape}
- \usepackage{colortbl}%
- \newcommand{\myrowcolour}{\rowcolor[gray]{0.925}}
- \usepackage{wasysym}
-
-    \makeatother
-
-\lstset{
-  language=Python,
-  showstringspaces=false,
-  formfeed=\newpage,
-  tabsize=4,
-  commentstyle=\itshape\color{light-gray},
-  basicstyle=\ttfamily\scriptsize,
-  morekeywords={lambda, self, assert, as, cls},
-  numbers=left,
-  numberstyle=\scriptsize\color{light-gray}\textsf,
-  xleftmargin=2em,
-  stringstyle=\color{mauve}
-}
-\lstdefinestyle{Bash}{
-    language={}, 
-    numbers=left,
-    numberstyle=\scriptsize\color{light-gray}\textsf,
-    moredelim=**[is][\color{blue}\bf\ttfamily]{`}{`},
-}
-\lstdefinestyle{OpenCL}{
-	language=C++,
-	morekeywords={kernel, __kernel, global, __global, size_t, get_global_id, sin, printf, int2}
-}
-
-\usepackage{float}
-\floatstyle{ruled}
-\newfloat{codelisting}{tp}{lop}
-\floatname{codelisting}{Listing}
-\setlength{\floatsep}{10pt}
-\setlength{\textfloatsep}{10pt}
-
-\usepackage{url}
-
-%\usepackage{todo}
-%\usepackage[subject={Todo},author={Josef}]{pdfcomment}
-%\usepackage{cooltooltips}
-\newcommand{\todo}[1]{{\color{red} #1}}
-\usepackage{placeins}
-
-\usepackage{textpos}
-
-\renewcommand\topfraction{0.85}
-\renewcommand\bottomfraction{0.85}
-\renewcommand\textfraction{0.1}
-\renewcommand\floatpagefraction{0.85}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\usepackage{tabularx}
+\usepackage{morefloats}
 \usepackage{bussproofs}    % Gentzen-style deduction trees *with aligned sequents!*
@@
 \usepackage{syntax}        % For type-setting formal grammars.
-
 \usepackage{hyperref}       % For links in citations
-
-\usepackage{float}			% Make figures float if [H] option is passed.
-
-\iffalse
-\usepackage{listings}		% For typesetting code listings
-\usepackage{callout}
-\usepackage{titlesec}
-\usepackage[T1]{fontenc}
-\usepackage{upquote}
-\lstset{upquote=true}
-\fi
-
-\usepackage{textcomp}		% For \textquotesingle as used in introduction
-\usepackage{color}			% for box colors, like in TAPL.
-
-\usepackage{amsmath}		% Begin Carthage default packages
-\usepackage{makeidx}
-\usepackage{amsthm}
-\usepackage{amsfonts}
-\usepackage{amssymb}
-\usepackage{graphics}
 \usepackage{enumerate}
-\usepackage{multicol}
-\usepackage{epsfig}
-\usepackage{csquotes}
 \usepackage{enumitem}
+
+\input{../../att-icfp14/macros-atlam}
+
+\usepackage{todonotes}
 
@@
 
-\title{Statically Typed String Sanitation Inside a Python: Technical Report}
-\author{Nathan Fulton\and Cyrus Omar\and Jonathan Aldrich}
-
+\title{Statically Typed String Sanitation Inside a Python (Technical Report)}
+\author{Nathan Fulton \and Cyrus Omar \and Jonathan Aldrich}
+\date{December 2014}
+\trnumber{CMU-ISR-14-112}
+\keywords{type systems; regular languages; input sanitation; string sanitation}
+\support{This work was supported by the National Security Agency lablet contract \#H98230-14-C-0140.}
+\abstract{This report contains supporting evidence for claims put forth and explained in the paper ``Statically Typed String Sanitation Inside a Python'' \cite{fulton2014}, including proofs of lemmas
+and theorems asserted in the paper, examples, and additional discussion of the paper's technical content.}
 \begin{document}
-\linenumbers
-
 \maketitle
-
-\begin{abstract}
-Web applications must ultimately command systems like web browsers and database engines using strings.
-Strings derived from improperly sanitized user input  can thus be a vector for command injection attacks.
-
-In this report, we introduce \emph{regular string types}, which classify strings known statically to be in a specified regular language.
-These types come equipped with common operations like concatenation, substitution and coercion, so they  can be used to implement, in a conventional manner, the portions of a web application or application framework that must directly construct command strings.
-Simple type annotations at key interfaces can be used to statically verify that sanitization has been performed correctly without introducing redundant run-time checks.
-We specify this type system in a minimal typed lambda calculus, $\lambda_{RS}$.
-
-We then specify a translation from $\lambda_{RS}$ to a language fragment containing only standard strings and regular expressions and prove that the correctness theorem for
-$\lambda_{RS}$ is preserved under translation.
-\end{abstract}
-
-\section{Introduction}
-
-Command injection vulnerabilities are among the most common and severe security vulnerabilities in modern web applications.
-They arise because web applications, at their boundaries, control external systems using commands represented as  strings.
-In \cite{fulton2014} the authors argue that extensible type systems are an attractive solution to this important security problem.
-This Technical Report contains supporting evidence for claims put forth and explained in \cite{fulton2014}, including proofs of lemmas
-and theorems asserted in the paper, examples, and additional discussion of the paper's technical content.
-
+\section{Terminology and Notation}
 Theorems and lemmas appearing in \cite{fulton2014} are numbered, while supporting facts appearing only in the Technical Report are lettered.
@@
 
-\section{Proofs of Lemmas and Theorems about $\lambdas$}
+\section{$\lambdas$}
 
@@
 \bibliographystyle{abbrv}
-
 \bibliography{research}
-
 \end{document}

### 2014-11-24 thesis-proposal/background.tex
Updated proposal

--- 2014-03-15 thesis-proposal/background.tex
@@
 % !TEX root = omar-thesis-proposal.tex
-\subsubsection{Background}\label{alibs}\label{background}
-The term \emph{active libraries} was first introduced by Veldhuizen et al.
+% \subsubsection{Background}\label{alibs}\label{background}
+% The term \emph{active libraries} was first introduced by Veldhuizen et al.
 \cite{activelibraries, active-libraries-thesis} to describe ``libraries that take an active role in compilation, rather than being passive collections of subroutines''.
-The authors suggested a number of reasons libraries might benefit from being able to influence the programming system directly, including high-level program optimization, checking programs for correctness against specialized criteria, reporting domain-specific errors and warnings, and ``rendering domain-specific textual and non-textual program representations and for interacting with such representations'' (anticipating interactions between libraries and tools other than just the compiler).
+The authors suggested a number of reasons libraries might benefit from being able to influence the programming system directly, including high-level program optimization, checking programs for correctness against specialized criteria, reporting domain-specific errors and warnings, and ``rendering domain-specific textual and non-textual program representations and for interacting with such representations''.%(anticipating interactions between libraries and tools other than just the compiler).
 
-% This stuff isn't actually hooking into the compiler...
-The first concrete realizations of active libraries in statically typed settings, prompting the introduction of the term, were libraries that performed domain-specific program optimization at compile-time by exploiting language mechanisms that allow for limited compile-time computation.
+% % This stuff isn't actually hooking into the compiler...
+% The first concrete realizations of active libraries in statically typed settings, prompting the introduction of the term, were libraries that performed domain-specific program optimization at compile-time by exploiting language mechanisms that allow for limited compile-time computation.
 A prominent example in the literature is Blitz++, a library that uses C++ template metaprogramming to optimize compound operations on vectors and matrices by eliminating intermediate allocations \cite{veldhuizen2000blitz++}.
@@
 
-More powerful and direct compile-time \emph{term rewriting mechanisms} available in some languages can also be used for optimization, as well as for introducing specialized error checking logic and implementing new abstractions.
-For example, typed macros, such as those in MetaML \cite{Sheard:1999:UMS}, Template Haskell \cite{SheardPeytonJones:Haskell-02} and Scala \cite{scala-macros}, take full control over all of the code that they enclose.
+% More powerful and direct compile-time \emph{term rewriting mechanisms} available in some languages can also be used for optimization, as well as for introducing specialized error checking logic and implementing new abstractions.
+For example, typed macros, such as those in MetaML \cite{Sheard:1999:UMS}, Template Haskell \cite{SheardPeytonJones:Haskell-02} and Scala \cite{ScalaMacros2013}, take full control over all of the code that they enclose.
 This can be problematic, however, as outer macros can interfere with inner macros.
@@
 Thus, macros represent at best a partial solution: they can be used to automate code generation, but not to globally extend the syntax or static guarantees of a language.
-It can also be difficult to reason about the semantics of code when any number of enclosing macros may be manipulating it, and to build tools that operate robustly in their presence.
+It can also be difficult to reason about the semantics of code when any number of enclosing macros may be manipulating it (particularly if they can change its type, as in some static macro systems), and to build tools that operate robustly in their presence.
 
-Some term rewriting systems replace the delimited scoping of macros with global pattern-based dispatch.
+% Some term rewriting systems replace the delimited scoping of macros with global pattern-based dispatch.
 Xroma (pronounced ``Chroma''), for example, allows users to insert custom rewriting passes into the compiler from within libraries \cite{activelibraries}.
@@
 It is also difficult to guarantee that such logic is correct and difficult to reason about code when simply importing a library can change the semantics of the program in a highly non-local manner.
-% http://www.haskell.org/ghc/docs/latest/html/users_guide/rewrite-rules.html
+% % http://www.haskell.org/ghc/docs/latest/html/users_guide/rewrite-rules.html
 
-Another example of an active library approach to extensibility with non-local scope is SugarJ \cite{erdweg2011sugarj} and other languages generated by Sugar* \cite{erdweg2013framework}, like SugarHaskell \cite{erdweg2012layout}.
+% Another example of an active library approach to extensibility with non-local scope is SugarJ \cite{erdweg2011sugarj} and other languages generated by Sugar* \cite{erdweg2013framework}, like SugarHaskell \cite{erdweg2012layout}.
 These languages permit libraries to extend the base syntax of the core language in a nearly arbitrary manner, and these extensions are imported transitively throughout a program.
@@
 For example, a library that defines a literal syntax for HTML would conflict with another that defines a literal syntax for XML because they define differing semantics for some of the same syntactic forms.
-If SugarJ was used by two different regular expression engines to provide literal syntax for regular expression patterns, there could easily be conflicts at link-time because both will introduce many of the same notations but back them with differing implementations.
-And again, it is difficult to predict what an unfamiliar piece of syntax desugars into, leading to difficulties reading and reasoning about code.
+If SugarJ was used by two different regular expression engines to provide literal syntax for regular expression patterns, there could easily be conflicts at link-time because both will introduce many of the same syntax but back them with differing implementations.
+% And again, it is difficult to predict what an unfamiliar piece of syntax desugars into, leading to difficulties reading and reasoning about code.
 
-%Another example comes from the Tau package for tuning and analyzing parallel programs, which allows libraries to instrument themselves declaratively so that internal details can be hidden from users of the tool \cite{tau}.
+% %Another example comes from the Tau package for tuning and analyzing parallel programs, which allows libraries to instrument themselves declaratively so that internal details can be hidden from users of the tool \cite{tau}.
 
-%The researchers behind these tools often claim that such conflicts are rare.
+% %The researchers behind these tools often claim that such conflicts are rare.
 There have, to our knowledge, been no studies of this claim.
@@
 
-%There also does not appear to be a clear, well-founded theoretical foundation for this approach, nor for active libraries in general.
+% %There also does not appear to be a clear, well-founded theoretical foundation for this approach, nor for active libraries in general.
 
-%\todo{Smooth this transition}Developers specify type-specific compile-time and edit-time features by equipping type families, when they are defined, with functions written in an appropriate metalanguage.
+% %\todo{Smooth this transition}Developers specify type-specific compile-time and edit-time features by equipping type families, when they are defined, with functions written in an appropriate metalanguage.
 These are selectively invoked by tools, such as the parser, type checker, translator and editor, when they work with expressions of that type, and only in those situations.
 
-%\subsection{Example: First-Class Natural Numbers}
-%To make our approach more concrete and emphasize its foundational character, let us begin with a very simple example, first-class natural numbers.
+% %\subsection{Example: First-Class Natural Numbers}
+% %To make our approach more concrete and emphasize its foundational character, let us begin with a very simple example, first-class natural numbers.
 If a programming system included first-class support for natural numbers, it might support behaviors like the following:
-%
-%\begin{enumerate}
-%\item type checking rules that constrain uses of the three operators associated with the $\nat$ type, as defined in G\"{o}del's System \textbf{T} \cite{tapl}: $\natz$, $\nats{e}$ and $\natrec{e_1}{e_2}{e_3}$\todo{phrasing of his}
-%\item type checker error messages that provide domain-specific assistance, like detecting when the trailing $\textbf{z}$ in a constant has been omitted: \todo{phrasing / merge this into 1?}
-%\begin{verbatim}
-%    s(s(s))
-%        ^ (Line 1, Character 5)
-%    Error: Operator s requires 1 argument of type nat. Provided 0 arguments.
-%    Did you mean s(z)?
-%\end{verbatim}
-%\item translation into an efficient integer representation at run-time
-%\item syntax support \todo{elaborate on syntax support for nat literals}
-%\item editor support, such as the ability to instantiate nat constants more easily \todo{elaborate on this / reference Graphite section}
-%%generate statistical predictions useful for code completion (e.g. biasing smaller constants) and other editor interactions, customized syntactic/visual representations and so on.
-%\end{enumerate}
-%
+% %
+% %\begin{enumerate}
+% %\item type checking rules that constrain uses of the three operators associated with the $\nat$ type, as defined in G\"{o}del's System \textbf{T} \cite{tapl}: $\natz$, $\nats{e}$ and $\natrec{e_1}{e_2}{e_3}$\todo{phrasing of his}
+% %\item type checker error messages that provide domain-specific assistance, like detecting when the trailing $\textbf{z}$ in a constant has been omitted: \todo{phrasing / merge this into 1?}
+% %\begin{verbatim}
+% %    s(s(s))
+% %        ^ (Line 1, Character 5)
+% %    Error: Operator s requires 1 argument of type nat. Provided 0 arguments.
+% %    Did you mean s(z)?
+% %\end{verbatim}
+% %\item translation into an efficient integer representation at run-time
+% %\item syntax support \todo{elaborate on syntax support for nat literals}
+% %\item editor support, such as the ability to instantiate nat constants more easily \todo{elaborate on this / reference Graphite section}
+% %%generate statistical predictions useful for code completion (e.g. biasing smaller constants) and other editor interactions, customized syntactic/visual representations and so on.
+% %\end{enumerate}
+% %
 
-%\section{Related Work}
-%We review existing approaches that are available to researchers and domain experts who wish to develop novel constructs below.
-%
-%\subsection{Language-Oriented Approaches}
-%\subsubsection{Language Frameworks}
-%A number of tools have been developed to assist with this task of developing new languages and their associated tools, including compiler generators, language workbenches and domain-specific language frameworks (see \cite{fowler2010domain} for a review).
+% %\section{Related Work}
+% %We review existing approaches that are available to researchers and domain experts who wish to develop novel constructs below.
+% %
+% %\subsection{Language-Oriented Approaches}
+% %\subsubsection{Language Frameworks}
+% %A number of tools have been developed to assist with this task of developing new languages and their associated tools, including compiler generators, language workbenches and domain-specific language frameworks (see \cite{fowler2010domain} for a review).
 In some cases, these tools allow language features to be defined modularly and composed differentially to produce a variety of different languages.
@@
 User-defined code is still ultimately compiled against a particular language, and because language composition cannot be automatic or guaranteed safe, interoperability with code written in other languages is thus limited by the difficulties described above despite the modular construction.
-%
-%\subsubsection{Extensible Compilers and Tools}
-%A related methodology is to implement language features as compiler and tool extensions directly.
+% %
+% %\subsubsection{Extensible Compilers and Tools}
+% %A related methodology is to implement language features as compiler and tool extensions directly.
 A number of extensible compilers have been developed to support this approach (see \cite{clements2008comparison}).
@@
 We argue that this approach should be considered harmful.
-%
-%\subsection{Library-Oriented Approaches}
-%\subsubsection{Embedded DSLs}
-%Embedded domain-specific languages are languages that creatively repurpose existing language constructs to create interfaces that resemble those of a distinct language.
+% %
+% %\subsection{Library-Oriented Approaches}
+% %\subsubsection{Embedded DSLs}
+% %Embedded domain-specific languages are languages that creatively repurpose existing language constructs to create interfaces that resemble those of a distinct language.
 In languages with rich type systems, such as Haskell, this approach can be quite successful (e.g.
@@
 Ultimately, however, this approach is limited by the host system, and as discussed in Section 2, thus limits experts who want to express particularly novel constructs, relative to the host language.
-%
-%\subsubsection{Term Rewriting Systems}
-%Many languages and tools allow developers to rewrite expressions according to custom rules.
+% %
+% %\subsubsection{Term Rewriting Systems}
+% %Many languages and tools allow developers to rewrite expressions according to custom rules.
 These can broadly be classified as {\it term rewriting systems}.
@@
 Finally, term rewriting systems focus on rewriting terms to support alternative language semantics but do not intrinsically support extensions that cover the full programming system.
-%
-%\subsubsection{Active Libraries}
-%Active libraries, as proposed by Czarnecki et al.
+% %
+% %\subsubsection{Active Libraries}
+% %Active libraries, as proposed by Czarnecki et al.
 \cite{activelibraries}, ``are not passive collections of routines or objects, as are traditional libraries, but take an active role in generating code''.
 They go on to suggest a number of areas in which the library could interact with the programming system, including optimizing code, checking source code for correctness, reporting domain-specific errors and warnings, and ``rendering domain-specific textual and non-textual program representations and for interacting with such representations''.
-%
-%This paper was largely a proposal.
+% %
+% %This paper was largely a proposal.
 The concrete implementations of this concept have largely been term rewriting systems described above.
@@
 
-%\subsection{Type-Level Computation} %Haskell, Ur and $\Omega$mega
-%System XX with simple case analysis provides the basis of type-level computation in Haskell (where type-level functions are called type families \cite{Chakravarty:2005:ATC}).
+% %\subsection{Type-Level Computation} %Haskell, Ur and $\Omega$mega
+% %System XX with simple case analysis provides the basis of type-level computation in Haskell (where type-level functions are called type families \cite{Chakravarty:2005:ATC}).
 Ur uses type-level records and names to support typesafe metaprogramming, with applications to web programming \cite{conf/pldi/Chlipala10}.
@@
 
-%\subsubsection{Run-Time Indirection}
-%{\it Operator overloading} \cite{vanWijngaarden:Mailloux:Peck:Koster:Sintzoff:Lindsey:Meertens:Fisker:acta:1975} and {\it metaobject dispatch} \cite{Kiczales91} are run-time protocols that translate operator invocations into function calls.
+% %\subsubsection{Run-Time Indirection}
+% %{\it Operator overloading} \cite{vanWijngaarden:Mailloux:Peck:Koster:Sintzoff:Lindsey:Meertens:Fisker:acta:1975} and {\it metaobject dispatch} \cite{Kiczales91} are run-time protocols that translate operator invocations into function calls.
 The function is typically selected according to the type or value of one or more operands.
@@
 
-%\subsection{Specification Languages}
-%Several {\it specification languages} (or {\it logical frameworks}) based on these theoretical formulations exist, including the OBJ family of languages (e.g.
+% %\subsection{Specification Languages}
+% %Several {\it specification languages} (or {\it logical frameworks}) based on these theoretical formulations exist, including the OBJ family of languages (e.g.
 CafeOBJ \cite{Diaconescu-Futatsugi01}).
@@
 The  language itself specifies which verification and execution strategies are used.
-%
-%Type-determined compilation takes a more concrete approach to the problem, focusing on combining {\it implementations} of different\- logics, rather than simply their specifications.
+% %
+% %Type-determined compilation takes a more concrete approach to the problem, focusing on combining {\it implementations} of different\- logics, rather than simply their specifications.
 In other words, it focuses on combining {\it type checkers} and {\it implementation strategies} rather than more abstract representations of a language's type system and dynamic semantics.
 In Section 4, we outlined a preliminary approach based on proof assistant available for the type-level language to unify these approaches, and we hope to continue this line of research in future work.
-%
+% %

### 2014-11-24 thesis-proposal/problem.tex
Updated proposal

@@
 \section{Motivation}\label{motivation}
-\begin{quote}\textit{The recent development of programming languages suggests that the simultaneous achievement of simplicity 
+\begin{quote}\textit{The recent development of programming languages suggests that the simul\-taneous achievement of simplicity 
 and generality in language design is a serious unsolved 
-problem.}
--- John Reynolds, 1970 \cite{Reynolds70}\end{quote}
+problem.} -- John Reynolds, 1970 \cite{Reynolds70}\end{quote}
 %%One might exp ``generality''
 %Designing a general-purpose programming language along with its supporting tools (collectively, a \emph{programming system}) has long stood as a grand challenge in computing. 
-Achieving both simplicity and generality in a programming language and its associated tools -- collectively, a \emph{programming system} -- requires organizing around a small number of mechanisms that support the realization of many others   as libraries.
+Achieving both simplicity and generality in the design of a programming language and its associated tools (collectively, a \emph{programming system}) requires organizing around a small number of mechanisms that support the expression of many others   as libraries.
 % In such a system, new dialects and revisions of the system would rarely be necessary.
@@
 %Not all useful abstractions can be  realized, when considered comprehensively, as libraries.% The problem of achieving both simplicity and generality simultaneously with a single system remains unsolved.
-Much progress has been made in the decades since Reynolds' statement above, but a programming system that comprehensively achieves these goals remains elusive, as evidenced by the fact that across all language lineages,  new language and tool {dialects} remain common vehicles for new features and abstractions, both in research and practice.
+Much progress has been made in the decades since Reynolds' statement above, but  across all language lineages,  new language and tool \emph{dialects} remain common vehicles for new ideas, both in research and practice.
+This suggests that library-based mechanisms are still not general enough to encompass all desirable modes of expression.
 %Library implementations are not yet satisfying.
 
-We observe that new system dialects are often motivated by a desire for new concrete syntax, new type system fragments, or new editor services, all {features} of programming systems that library providers usually do not have substantial control over.
+We will consider the broad class of situations where providers need to introduce new concrete syntax, type system fragments, or editor services into a system to realize an idea.
 %, encouraged  historically  by the availability of tools like compiler generators and,  more recently, language workbenches \cite{workbenches} and DSL frameworks \cite{dsl}.
 Unfortunately, taking this approach makes it substantially more difficult for clients to import high-level abstractions orthogonally.
-Introducing mechanisms that decentralize control over these features could decrease the need for new dialects and thus increase generality.
-But this must done with care, because giving away too much control could weaken important metatheoretic guarantees.
-Moreover, such  mechanisms must come equipped with \emph{modular reasoning principles} for it to be practical for library clients to rely upon them.
-Having to manually resolve conflicts between libraries or reestablish important properties of a program whenever a new library is imported introduces undesirable complexity.
+Equipping the system with \emph{extension mechanisms} that decentralize control over these features would  decrease the need for dialects, but this must done with care, because giving away too much control could also weaken the metatheory of the system.
+For example, were extensions allowed to arbitrarily modify the type system of the language, there would be immediate questions about {type safety}.
+Moreover, extension   mechanisms must come with \emph{modular reasoning principles} for it to be practical for library clients to rely upon them.
+Having to resolve conflicts between extensions or reestablish correctness properties of a program whenever a new extension is imported is unacceptably complex.
 %It can also make it more difficult to define new sorts of tools for working with the language because they cannot rely on there being a fixed abstract syntax (the canonical \emph{expression problem}).
@@
 
-In this thesis, we aim to design mechanisms that handle these issues by embracing the already widely accepted notion that \emph{type constructors} (or more concisely, \emph{tycons}) can serve as the fundamental organizational unit of a programming system.
-We show how to delegate control over certain aspects of the concrete syntax, external type system and editor services to user-defined functions associated with type constructors, together forming what we call \emph{active type constructors}.
-This permits library-based realizations of features that today would need to be built into a system by its central designers.
-By layering these mechanisms over a fixed typed internal language and constraining these functions appropriately, we retain key  metatheoretic properties and arrive at powerful modular reasoning principles.
+In this thesis, we will design extension mechanisms that handle these issues, permitting library-based expression of features that today would need to be built into a system by its central designers.
+More specifically, we show how to delegate control over certain aspects of the concrete syntax, external type system and editor services to \emph{static functions}, i.e.
+user-defined functions written in a \emph{static language}.
+By layering these mechanisms over a fixed typed internal language, constraining the static language appropriately and validating the code it generates, we will retain key  metatheoretic properties and arrive at powerful modular reasoning principles.
+%By associating these static functions with type constructors, forming \emph{active type constructors},  we will further show how they can enjoy a usage profile essentially identical to that of built in features.
 % while maintaining key metatheoretic properties and modular reasoning principles.
@@
 %To realize a new abstraction or system behavior, such experts can consider either a \emph{language-internal approach}, where they work within an existing language and distribute their solutions as libraries, or a \emph{language-external approach}, where they create a new, distinct programming system (often centered around what has come to be called a new \emph{domain-specific language} \cite{dsl}) or extend an existing system by some mechanism that is not part of the language itself, such as an extension mechanism supported by a {particular} compiler, editor or other tool.
-\subsection{Motivating Example: Regular Expressions}\label{regex}
-To make the use cases we seek to address more concrete, let us begin with a simple example that we will return to throughout this proposal.
+
+\subsection{Motivating Examples}\label{sec:examples}
+To make the use cases we seek to address more concrete, let us begin with a few simple examples that we will return to throughout this work.
+
+\subsubsection{Example 1: Regular Expressions}\label{sec:regex}
 \emph{Regular expressions} are commonly used to capture patterns in strings (e.g.
@@
 %The cognitive load of reading and writing patterns is low and patterns are parsed once at compile-time. 
-Malformed patterns would result in intelligible {parser} errors.
-\item A \textbf{type system} that ensures that key invariants related to regular expressions are statically maintained. It might statically ensure that: 
+Malformed patterns would ideally result in intelligible \emph{compile-time} parse errors.
+\item A \textbf{static semantics} that ensures that key invariants related to regular expressions are statically maintained. For example, 
 	\begin{enumerate}
-	\item only patterns, or properly escaped strings, are spliced into a pattern literal, to avoid splicing errors and injection attacks \cite{owasp2013, Bravenboer:2007:PIA:1289971.1289975}
-	\item out-of-bounds backreferences to a captured group do not occur \cite{spishak2012type}
-	\item string processing operations do not lead to a string that is malformed, when well-formedness can be captured as membership in a regular language \cite{fulton-thesis,HosoyaVouillonPierce2000ICFP}
+	\item At minimum, we would want that only patterns, or properly escaped strings, are spliced into a pattern literal like that above, to avoid splicing errors and injection attacks \cite{owasp2013, Bravenboer:2007:PIA:1289971.1289975}.
+For example,
+\begin{lstlisting}[numbers=none]
+let fourChars : string = '\d\d'
+let twoDigits : Pattern = <SURL\d\dEURL>
+let p : Pattern = <SURL{EURLfourCharsSURL}-{EURLtwoDigitsSURL}EURL>
+\end{lstlisting}
+    should either be ill-typed because \lstinline{fourChars} is not a pattern, or be equivalent to the following:
+\begin{lstlisting}[numbers=none]
+let p : Pattern = <SURL\\d\\d-\d\dEURL>
+\end{lstlisting}
+    It should \emph{not} treat the string incorrectly as a pattern:
+\begin{lstlisting}[numbers=none]
+let p : Pattern = <SURL\d\d-\d\dEURL> (* injection attack vector! *)
+\end{lstlisting}
+	\item A more advanced semantics might ensure that out-of-bounds backreferences to a captured group do not occur \cite{spishak2012type}.
+For example, we would want \verb|BisI|, above, to have the more precise type \verb|Pattern[1]|, i.e.
+a pattern with one captured group.
 	\end{enumerate}
@@
 %By performing this step ahead-of-time, these dangers can be avoided.
-\item \textbf{Editor services} to support syntax highlighting, consulting relevant documentation, interactive testing and pattern extraction from example strings.
+\item A range of \textbf{editor services} that support syntax highlighting for patterns, retrieval of relevant documentation, interactive testing and pattern extraction from example strings have been shown to be helpful for programmers when working with complex regular expressions \cite{ACC_VLHCC}.
+An example of some of these is shown in Figure \ref{fig:regex-palette}: the editor service shown helps programmers generate correct code, here via the Java standard library's implementation of regular expressions (discussed below).
 \end{enumerate}
+
+\begin{figure}
+\begin{center}
+\includegraphics[scale=.55]{regex-palette.pdf}\\
+$\Downarrow$ \text{(pressing \textbf{Enter})}\\
+\includegraphics[scale=.55]{regex-code-generated.pdf}
+\end{center}
+\vspace{-20px}
+\caption{An example of type-specific editor services, here shown for Java.}
+\label{fig:regex-palette}
+%\vspace{-10px}
+\end{figure}
 
@@
 %Unfortunately, it is impossible to completely define the syntax and the semantics referenced above using contemporary mechanisms, so library providers need to compromise.
+The most common strategy, exemplified by Java, is to ask clients to introduce patterns as strings, deferring their  parsing and compilation to run-time.
+This provides only a partial approximation to feature 1 due to clashes between string escape sequences and pattern escape sequences, as can be seen in the generated code at the bottom of Figure \ref{fig:regex-palette} (double backslashes are needed).
+Parsing errors due to these clashes have been observed to be difficult for even experienced programmers to diagnose \cite{Omar:2012:ACC:2337223.2337324}.
+Moreover, the static guarantees cannot be provided, leading to run-time exceptions (common even in well-tested code using regular expressions \cite{spishak2012type}), and mistakes that are not caught even at run-time, some of which can lead to security vulnerabilities due to injection attacks \cite{owasp2013}.
+It also introduces performance overhead due to run-time parsing and compilation of patterns, and redundant well-formedness checks.
 
-The most common strategy is to ask clients to introduce patterns as strings, deferring their  parsing, compilation and use to run-time.
-This provides only a partial approximation to feature 1 due to clashes between string escape sequences and regular expressions (parsing errors due to these clashes have been observed to be difficult for even experienced programmers to diagnose \cite{Omar:2012:ACC:2337223.2337324}).
-Moreover, none of the static guarantees can be provided, leading to run-time exceptions (common even in well-tested code \cite{spishak2012type}), and mistakes that are not caught even at run-time, some of which can lead to security vulnerabilities due to injection attacks \cite{owasp2013}.
-It also introduces performance overhead due to run-time parsing and compilation of patterns, and redundant run-time checks.
-Staging these computations can shift some of the errors and cost to compile-time, albeit with some additional semantic and syntactic complexity, but the issues related to the fact that a regular expression is not a string (e.g.
+%Staging these computations can shift some of the errors and cost to compile-time, albeit with some additional semantic and syntactic complexity, but the issues related to the fact that a regular expression is not a string (e.g.
 syntactic clashes and injection attacks) persist.% parsing can move some errors to run-time, albeit at additional syntactic cost, but it also does not straightforwardly handle splicing-related issues, as we will discuss later.
 
-A more semantically justifiable approach is to dispense with string representations entirely and directly work with an encoding of regular expression patterns using  general-purpose constructs like sums and  products (e.g.
-as exposed by functional datatypes or objects), or using an abstract type.
+A more semantically justifiable approach is to dispense with string representations entirely and directly work with an encoding of regular expression patterns using  general-purpose constructs like recursive sums and  products (e.g.
+as exposed by  datatypes in a functional language, shown in Sec.
+\ref{sec:syntax}, or a class hierarchy in an object-oriented language), or using an abstract data type.
 This provides feature 2a while sacrificing feature 1 (such an encoding affords only an approximation of the verbose abstract syntax of patterns, not their concrete syntax).
-Advanced invariants like those described in features 2b and 2c are not directly tracked by such encodings and require solutions beyond those commonly found in modern simply-typed languages (we will return to this  in Sec.
-\ref{att}).%Explicitly constructing a regular expression by applying datatype constructors is unacceptably verbose.
+Advanced invariants like those described in features 2b are not directly tracked by such encodings and require solutions beyond those commonly found in simply-typed languages (a feature like GADTs might suffice in this case \cite{XiCheChe03}).%Explicitly constructing a regular expression by applying datatype constructors is unacceptably verbose.
 
@@
 because they require the programmer to switch contexts and cannot make use of the code context \cite{Omar:2012:ACC:2337223.2337324}.
-Working with high-level abstractions like these in a ``low-level'' editor is thus awkward and error-prone, just like attempting to work with high-level abstractions directly in terms of an encoding.
-For these reasons,   we see editor services as within the scope of an abstraction specification.
+%Working with high-level abstractions like these in a ``low-level'' editor is thus awkward and error-prone, just like attempting to work with high-level abstractions directly in terms of an encoding.
+For these reasons,   we see editor services as within the broad scope of an abstraction specification.
+
+\subsubsection{Example 2: Regular Strings}\label{sec:rstr}
+The examples above all dealt with how regular expression patterns are represented.
+However, programmers also want to reason about which regular language a string value is in \cite{sanitation-psp14}.
+For example, we might want to ensure that the arguments to a function called \verb|connect| corresponding to a database username and password are only alphanumeric strings.
+We might include this information directly in the type of \verb|connect|, i.e.
+\begin{lstlisting}[numbers=none]
+type alphanumeric = rstring[<SURL[A-Za-z0-9]+EURL>]
+val connect : (alphanumeric * alphanumeric) -> DBConnection
+\end{lstlisting}
+
+This example uses a specialized \emph{type constructor}, \verb|rstring|, indexed by a statically known regular expression.
+The operations on such \emph{regular strings} would need a static semantics that track the regular language a string is in.
+For example, a regular string introduced by concatenating two other alphanumeric strings  should also be an alphanumeric string.
+
+Ideally, we would be able to use standard string literal syntax with regular strings, e.g. 
+\begin{lstlisting}[numbers=none]
+let connection = connect("admin", "password")
+\end{lstlisting}
+
+Regular strings could go beyond simply refining standard string types. For example, they might define operators specific to regular strings like \emph{captured group projection}:
+\begin{lstlisting}[numbers=none]
+let example : rstring[<SURL(\d\d\d)-(\d\d\d\d)EURL>] = "555-5555"
+let group0 (* : rstring[<\d\d\d>] *) = example#0
+\end{lstlisting}
+Ideally, it would be possible to define this operator such that its cost is $\mathcal{O}(1)$.
+%This example requires introducing new operators into the system that did not already exist, and using an underlying represention that differs for different type indices.
+
+\subsubsection{Example 3: Labeled Products with Functional Update Operators}\label{sec:lprod}
+The simplest semantics for product types is to define only nullary and binary products \cite{pfpl}.
+However, in practice, many variations on product types are built in to various dialects of ML and other languages: $n$-ary tuples, labeled tuples,
+records (identified up to reordering), and 
+records with width and depth coercions \cite{Cardelli:1984:SMI:1096.1098} and functional update operators \cite{ocaml-manual}.
+{The Haskell wiki notes that ''No, extensible records [i.e.
+records with a functional update operator] are not implemented in GHC.
+The problem is that the record design space is large, and seems to lack local optima.
+[...] As a result, nothing much happens.'' \cite{GHCFAQ}}%,
+ %mutable fields \cite{ocaml-manual}, 
+%field delegation \cite{atlang-gpce14} \todo{gpce submission} 
+%and 
+% ``methods'' (i.e. pure objects) \cite{TSLs}
+
+We would ideally like to avoid needing the language designer to decide \emph{a priori} on just one point in this large design space.
+Instead, they should simply provide some minimal way of defining products, perhaps only nullary and binary products, while making it possible for these other variations on products to be defined as libraries and used together without conflict.
+For example, we would like to be able to define labeled product types, which are like record types but maintain a row ordering.
+An example of a labeled product types classifying conference papers might be:
+\begin{lstlisting}[numbers=none]
+type Paper = lprod[{
+  title : rstring[<SURL.+EURL>], 
+   conf : rstring[<SURL[A-Z]+ \d\d\d\dEURL>]
+}]
+\end{lstlisting}
+
+Because of the row ordering, it should be possible to introduce values of this type with or without explicit labels, e.g.
+\begin{lstlisting}[numbers=none]
+fun make_paper(t : rstring[<SURL.+EURL>]) : Paper = {title=t, conf="EXMPL 2015"}
+\end{lstlisting}
+should be equivalent to
+\begin{lstlisting}[numbers=none]
+fun make_paper(t : rstring[<SURL.+EURL>]) : Paper = (t, "EXMPL 2015")
+\end{lstlisting}
+Note above that we aim to be able to re-use the standard record and tuple syntax.
+
+We should then be able to project out rows by providing a label (or a numeric index, not shown):
+\begin{lstlisting}[numbers=none]
+let test_paper = make_paper "Test Paper"
+test_paper#conf
+\end{lstlisting}
+
+We might also be wish to support functional update. For example, an operation that dropped a row might be used like this:
+\begin{lstlisting}[numbers=none]
+let title_only (* : lprod[{title : rstring[<.+>]}] *) = test_paper.drop[conf]
+\end{lstlisting}
+An operation that added a row, or updated the value of an existing row, might be used like this:
+\begin{lstlisting}[numbers=none]
+fun with_author(p : Paper, a : rstring[<.+>]) = p.ext(author=a)
+\end{lstlisting}
 \subsection{Language-External Approaches}\label{external-approaches}
-\begin{figure}
-\begin{center}
-\includegraphics[scale=.48]{approaches.pdf}
-\end{center}
-\vspace{-20px}
-\caption{\small (a) When taking a language-external approach, new features are packaged together into separate languages and tools.
-(b) When taking a language-integrated approach, there is one extensible host language and the compile-time and edit-time logic governing new constructs is expressed within ``active libraries''.}
-\label{approaches}
-%\vspace{-10px}
-\end{figure}
 
-In situations, like these, where a library-based approach is not satisfying, providers must take a \emph{language-external approach}, either by developing a new system dialect (supported by \emph{compiler generators} \cite{brooker1963compiler}, \emph{language workbenches} \cite{erdweg2013state}, \emph{DSL frameworks} \cite{fowler2010domain}, or simply by forking an existing codebase), or by using an extension mechanism for a {particular} compiler\footnote{Compilers that modify, or allow modification of, the semantics of their base language, rather than simply permitting semantics-preserving optimizations, should be considered a pernicious means for creating new language dialects.
+In situations, like these, where a library-based approach is not satisfying, providers must today take a \emph{language-external approach}, either by specifying a new system dialect (implementing it using a \emph{compiler generator} \cite{brooker1963compiler}, \emph{language workbench} \cite{erdweg2013state}, \emph{DSL framework} \cite{fowler2010domain}, or simply by forking an existing codebase), or by using an extension mechanism for a {particular} compiler\footnote{Compilers that modify, or allow modification of, the semantics of their base language, rather than simply permitting semantics-preserving optimizations, should be considered a pernicious means for creating new language dialects.
 That is, some programs that purport to be written in C, Haskell or Standard ML are actually written in compiler-specific dialects of these languages.}, editor or other tool.
@@
 A third researcher, developing a type system for reasoning about units of measure (\texttt{Q}) might again do the same.
-This results in a collection of distinct systems, as diagrammed in Figure \ref{approaches}a.
+This results in a collection of distinct programming systems, as diagrammed in Figure \ref{approaches}a.
+
 Although often justified as ``proofs of concept'', this is a rather weak notion because no abstraction will be used entirely in isolation and such a construction gives us no rigorously justifiable  reason to believe that it can safely and naturally coexist with others specified or implemented in the same way.
@@
 
-Recent evidence indicates that this is one of the major barriers preventing research on programming languages from being driven into practice.
-For example, developers prefer high-level language-integrated parallel programming abstractions that provide  strong semantic guarantees when all else is equal \cite{cave2010comparing}, but library-based approximations are far more widely adopted because  ``parallel programming languages'' privilege only a few chosen  abstractions at the language level.
-This is problematic because different abstractions are seen as more appropriate in different situations \cite{Tasharofi:2013rc}.
-Moreover,  parallel programming is rarely the only concern relevant to clients outside of a classroom or research setting.
-Support for regular expressions, for example, would be simultaneously desirable for processing large amounts of genomic data in parallel, but using these features together in the same compilation unit would be difficult or impossible if implemented using language-external means.
+Recent evidence indicates that this is one of the major barriers preventing research  from being driven into practice.
+For example, developers prefer high-level language-integrated parallel programming abstractions that provide  stronger semantic guarantees  \cite{cave2010comparing}, but library-based approximations are far more widely adopted because  ``parallel programming languages'' privilege only a few  abstractions at the language level.
+In contrast, it is widely acknowledged that  different abstractions are more appropriate in different situations \cite{Tasharofi:2013rc}.
+Moreover,  parallel programming is rarely the only relevant concern outside of a classroom or research setting.
+Support for regular expressions as above would be simultaneously desirable for processing large amounts of genomic data.% but using these features together in the same compilation unit would be difficult or impossible if implemented using language-external means.
 Indeed, switching to a ``parallel programming language'' would likely make it \emph{more} difficult to use regular expressions, as these are likely to be less well-developed in a specialized language than in an established general-purpose language.% This intuition was perhaps most succinctly expressed by a participant in a recent study by Basili et al.
@@
 
-\paragraph{Interoperability} Even in cases where, for each component of a software system, a programming system considered entirely satisfactory by its developers is available (e.g.
+% \paragraph{Interoperability} Even in cases where, for each component of a software system, a programming system considered entirely satisfactory by its developers is available (e.g.
 a team goes through the trouble of implementing \verb|G+R+P| by reading papers about \verb|G+R| and \verb|G+P| and disentangling orthogonality-related issues), there remains a problem at any interface between  components written using a different combination of features.
@@
 
-One strategy taken by proponents of a {language-oriented approach} \cite{journals/stp/Ward94} to partially address the interoperability problem is to  target an established intermediate language and use its constructs as a common language for communication between components written in different languages.
+% One strategy taken by proponents of a {language-oriented approach} \cite{journals/stp/Ward94} to partially address the interoperability problem is to  target an established intermediate language and use its constructs as a common language for communication between components written in different languages.
 Scala \cite{200464/IC} and F\# \cite{pickering2007foundations} are examples of prominent general-purpose languages that have taken this approach, and most DSL frameworks also rely on this strategy.
@@
 
-%This approach only works well when new languages consist of constructs that can also be expressed safely and almost as naturally in the common language.
-%But many of the most innovative constructs found in modern languages (often, those that justify their creation) are difficult to define in terms of existing constructs in ways that guarantee all necessary invariants are statically maintained and that do not require large amounts boilerplate code and run-time overhead.
-As a simple example with significant contemporary implications, F\#'s type system does not admit \verb|null| as a value for any type both defined and used within F\# code, but maintaining this sensible internal invariant still requires dynamic  checks because the stricter typing rules of F\# do not apply when F\# data structures are constructed by other languages on the Common Language Infrastructure (CLI) like C\# or SML.NET.
+% %This approach only works well when new languages consist of constructs that can also be expressed safely and almost as naturally in the common language.
+% %But many of the most innovative constructs found in modern languages (often, those that justify their creation) are difficult to define in terms of existing constructs in ways that guarantee all necessary invariants are statically maintained and that do not require large amounts boilerplate code and run-time overhead.
+% As a simple example with significant contemporary implications, F\#'s type system does not admit \verb|null| as a value for any type both defined and used within F\# code, but maintaining this sensible internal invariant still requires dynamic  checks because the stricter typing rules of F\# do not apply when F\# data structures are constructed by other languages on the Common Language Infrastructure (CLI) like C\# or SML.NET.
 This is not an issue exclusive to intermediate languages that make regrettable choices regarding \verb|null|, however.
@@
 \subsection{Language-Integrated Approaches}\label{language-integrated-approaches}
-We argue that, due to these problems with orthogonality and interoperability, taking a language-external approach to realizing a new feature should be considered harmful and avoided whenever possible.
-The goal of the research being proposed here is to design \emph{language-integrated extension mechanisms} that give providers the ability to define, within libraries, new features that, like those above, have previously required central planning,
+\begin{figure}
+\begin{center}
+\includegraphics[scale=.48]{approaches.pdf}
+\end{center}
+\vspace{-20px}
+\caption{\small (a) When taking a language-external approach, new features are packaged together into separate languages and tools.
+(b) When taking a language-integrated approach, there is one extensible host language and the compile-time and edit-time logic governing new constructs is expressed within ``active libraries''.}
+\label{approaches}
+%\vspace{-10px}
+\end{figure}
+We argue that, due to these problems, taking a language-external approach to realizing a new feature should be considered harmful and avoided whenever possible.
+The goal of the research being proposed here is to design \emph{language-integrated extension mechanisms} that give providers the ability to define, within libraries, new features that, like those above, that have previously required central planning,
 %\footnote{One might compare today's programming systems to  {centrally-planned} economies, whereas extensible\- systems more closely resemble modern market economies.
@@
 so that language-external approaches are less frequently necessary, as illustrated in Figure \ref{approaches}b. 
-Such libraries have been called \emph{active libraries}  \cite{activelibraries} in that they are not passive clients of features already available in the system.
-Features implemented within active libraries can be imported individually, unlike features implemented by external means, giving us a potential means to avoid the problems of orthogonality and interoperability.
+Such libraries have been called \emph{active libraries}  \cite{activelibraries} in that they more actively influence the syntax and semantics of the system.
+%Features implemented within active libraries can be imported individually, unlike features implemented by external means, giving us a potential means to avoid the problems of orthogonality and interoperability just described.
 
@@
 %Resolving these kinds of ambiguities requires significantly more expertise with parser technology than using the syntax itself does.
-These issues have plagued previous attempts to design language-integrated extensibility mechanisms.
-We will briefly review some of these attempts below, then return to our approach.% To prevent them, our mechanisms will organize  extension logic around types to guarantee that extensions are both safe in isolation and also safely composable in any combination.
+These issues have plagued previous attempts to design language-integrated extensibility mechanisms.\todo{I can include a more detailed related work section if requested.}% We will briefly review some of these attempts below, then return to our approach.% To prevent them, our mechanisms will organize  extension logic around types to guarantee that extensions are both safe in isolation and also safely composable in any combination.
 

### 2014-11-25 att-esop15/att-esop15.tex
Updated proposal

@@
 
+\usepackage{times}
 \renewcommand{\ttdefault}{txtt}
@@
 \sloppy
+
+\setlength{\abovecaptionskip}{5px}
+\setlength{\belowcaptionskip}{5px}
+
+\newcommand{\moutput}{^{\color{gray}+}}
+\newcommand{\rulename}[1]{(#1)}
+\def \TirNameStyle #1{\small\rulename{#1}}
+\renewcommand{\MathparLineskip}{\lineskiplimit=.6\baselineskip\lineskip=.6\baselineskip plus .2\baselineskip}
 
@@
 \begin{abstract}
-Abstraction providers sometimes need to extend a programming language with new base types and their corresponding operators. 
-Ideally, such language fragments could be defined and reasoned about separately.
-Unfortunately, existing tools and techniques do not come equipped with strong modular reasoning principle -- fundamentally, %comparable to those available for library-based embeddings packaged using a modern module system.
-each combination of fragments becomes its own dialect of the language for which metatheoretic and compiler correctness results must be established monolithically. 
-Recent work has started to address this problem, e.g.
+Researchers commonly describe typed programming languages in fragments or as simple calculi, leaving to language designers the task of composing these to form complete languages.
+%comparable to those available for library-based embeddings packaged using a modern module system.
+Metatheoretic results must be established monolithically for each such language, guided only informally by the results derived for simpler calculi.
+We argue that, as the language design space grows, mechanisms for defining and composing typed language fragments that provide strong modular reasoning principles are needed.%, so that important metatheoretic results need only be established locally and, ultimately, so that language extensions can be treated like libraries.
+%,  Sometimes this fails because certain language features do not ``play well'' together, but precisely characterizing which is elusive given these practices.
+
+%Recent work has started to address this problem, e.g.
 by showing how to modularly reason about desugarings atop a fixed type system.
@@
 the expression problem)  by delegating control over a small, uniform abstract syntax in a type-directed manner to logic associated with a type constructor.
-The paper is organized around a minimal calculus, @$\lambda$, that permits surprisingly practical examples.
-We establish several strong semantic guarantees, notably \emph{type safety}, \emph{stability of typing} under extension and \emph{conservativity}:
-that the \emph{type invariants} that a finite set of fragments maintain are conserved under  extension. 
-This involves lifting typed compilation techniques into the semantics and enforcing an abstraction barrier around each fragment using a form of ``internal'' type abstraction. 
-Conservativity then follows from classical parametricity results, so these \emph{modular type constructors} can be reasoned about like modules in an ML-like language. 
+
+In this paper, we begin from first principles with a minimal calculus, @$\lambda$, specified as a bidirectionally typed translation semantics: external terms are simultaneously typechecked and given a translation to a fixed typed internal language.
+The external language is made extensible by indexing the typing judgement by a \emph{tycon context}.
+Each tycon defines the semantics of its associated operators using a fixed static language where types are values.
+This language is constructed to ensure that several strong semantic guarantees follow from reasoning that can be performed in the ``closed world'' of a fixed tycon context, notably \emph{type safety} and \emph{conservativity}: that the \emph{type invariants} maintained under a minimal tycon context will be conserved under any further extended context.
+Violations are caught during typechecking by lifting typed compilation techniques into the semantics and enforcing an abstraction barrier around each tycon using type abstraction, so these \emph{modular type constructors} can be reasoned about modularly, i.e.
+like modules in an ML-like language.
 %\keywords{extensible languages; typed compilation; type-level computation; type abstraction}
@@
 \section{Introduction}
-Typed programming languages are often described in \emph{fragments}, each defining contributions to a language's concrete syntax, abstract syntax, static semantics and dynamic semantics. 
-In his textbook, Harper organizes fragments around type constructors, each introduced in a different chapter \cite{pfpl}.
-A language is then identified by a set of type constructors, e.g.
+Typed programming languages are often described in \emph{fragments}, each defining distinct contributions to a language's concrete and abstract syntax, static semantics and dynamic semantics. 
+For example, in his textbook, Harper organizes fragments around type constructors, describing each in a different chapter \cite{pfpl}.
+Complete languages are then identified by a set of type constructors, e.g.
 $\mathcal{L}\{{\rightharpoonup}\,{\forall}\,{\mu}\,{\keyw{1}}\,{\times}\,{+}\}$ is the language that builds in partial function types,  polymorphic types, recursive types, nullary and binary product types and binary sum types (its syntax is shown in Figure \ref{syntax-IL}, discussed further below).
-It is left implied that the metatheory developed separately is conserved when fragments like these are composed to form a language.
+Another common pattern is for a researcher to describe a new fragment by defining a complete calculus, perhaps including a ``catch-all'' constant and base type to stand for all other terms and types that may be of interest.
 
-Luckily, fragment composition is not an everyday programming task, because fragments like these are ``general purpose'' in that they make it possible to construct \emph{isomorphic embeddings} of many other fragments as ``libraries''.
-For example, lists can be placed in isomorphism with polymorphic recursive sum of products, $\iforall{\alpha}{\imu{t}{\iunit + (\alpha \times t))}}$.
-Languages providing datatypes in the style of ML  are perhaps most directly oriented around embeddings like this (preserving type disequality requires adding some form of generativity, as ML datatypes also expose).
+Metatheoretic reasoning techniques for programming languages generally assume that a complete specification of a language is being considered, however.
+To use these techniques, every \emph{combination} of fragments must be treated as its own monolithic language for which metatheorems must be established anew, guided only informally by results derived for the smaller systems from which they are, notionally, composed.
+
+%In both cases, it is left as an extrinsic concern to ensure that the metatheory developed separately is actually conserved when fragments defined in these ways are composed to form a complete language.
+
+Luckily, fragment composition is not an everyday programming task because fragments like these are ``general purpose'' in that they make it possible to construct \emph{isomorphic embeddings} of many other fragments as ``libraries''.
+For example, lists need not be built in ``primitively'' because they can be placed in isomorphism with the polymorphic recursive sum of products $\iforall{\alpha}{\imu{t}{\iunit + (\alpha \times t))}}$.
+Languages providing datatypes al\'a ML  are perhaps most directly oriented around such embeddings.
 % Leaving the core language simple makes it easier to establish its metatheory and verify that tools, like compilers, are implemented correctly.% They also have encouraging connections to logic.
@@
 
-Unfortunately, situations do sometimes arise where using these fragments to establish an isomorphic embedding that preserves a desirable fragment's  static and dynamic semantics (including bounds specified by its cost semantics) is not possible.
-Embeddings can also sometimes be unsatisfyingly \emph{complex}, as measured by the cost of the extralinguistic computations that are needed to map in and out of the embedding and, if these  must be performed mentally, considering  cognitive metrics like  error message comprehensibility.
+Although these fragments certainly appear to occupy a ``sweet spot'' in the design space, ``general purpose'' does not mean ``all-purpose'' and situations do continue to arise where using these fragments to establish an isomorphic embedding that preserves a desirable fragment's  static and dynamic semantics (including bounds specified by a cost semantics) is not possible.
+Embeddings can also sometimes be unsatisfyingly \emph{complex}, as measured by the cost of the extralinguistic computations that are needed to map in and out of the embedding and, if these  must be performed mentally by programmers, considering  cognitive factors.
 %We will discuss specific examples below.
-
-When an embedding is too complex, abstraction providers have a few options, discussed in Sec.
-\ref{prior-work}.
-When an embedding is not possible, however, or when these options are  insufficient, providers are often compelled to introduce these fragments by extending an existing language, thereby forming a new \emph{dialect}.
+%When an embedding is too complex, abstraction providers have a few options, discussed in Sec. \ref{prior-work}. When an embedding is not possible, however, or when these options are  insufficient, 
+%providers are often compelled to introduce these fragments by extending an existing language, thereby forming a new \emph{dialect}.
 To save effort, they may do so by forking existing artifacts or leverage tools like compiler generators or language frameworks, also reviewed in Sec.
@@
 %Indeed, the proliferation of language dialects constructed for this reason might be taken as an evidence that the core language is not as ``general'', when considered comprehensively, as might be hoped.
-Within the ML lineage, dialects that go  beyond $\mathcal{L}\{{\rightharpoonup}\,{\forall}\,{\mu}\,{\keyw{1}}\,{\times}\,{+}\}$ abound:
+Each time a researcher seeks to address one of these problems by designing a new language construct, a new \emph{dialect} of the language is born.
+Within the ML lineage, dialects that go  beyond core ML abound:
 %Reynolds, in a remark that recalls the ``Turing tarpit'' of Perlis \cite{Perl82a}, summarizes the issue \cite{Reynolds94anintroduction}: 
@@
 \begin{enumerate}
-\compresslist
-\vspace{5px}
+%\compresslist
 \item 
@@
 ``methods'' (i.e.
-pure objects \cite{conf/oopsla/Aldrich13, TSLs}).\footnote{The Haskell wiki notes that ''No, extensible records are not implemented in GHC.
+pure objects \cite{TSLs}).\footnote{The Haskell wiki notes that ''No, extensible records are not implemented in GHC.
 The problem is that the record design space is large, and seems to lack local optima.
-And all reasonable variants break backward compatibility.
-As a result, nothing much happens.'' \cite{GHCFAQ}}
- Sum types are also exposed in various ways: 
+[...] As a result, nothing much happens.'' \cite{GHCFAQ}}
+ Sum-like types are also exposed in various ways: 
 finite datatypes, 
@@
 \textbf{Specialized Fragments:} Fragments that track specialized static invariants to provide stronger correctness guarantees, manage unwieldy lower-level abstractions or control cost are also often introduced in dialects, e.g.
-for data parallelism  \cite{chakravarty2007data}, distributed programming \cite{Murphy:2007:TDP:1793574.1793585}, reactive programming \cite{mandel2005reactiveml}, authenticated data structures \cite{Miller:2014:ADS:2535838.2535851}, databases \cite{Ohori:2011:MSM:2034773.2034815} and units of measure \cite{conf/cefp/Kennedy09}.% All of these are implemented as dialects of existing languages, presumably because a strong encoding was not feasible.
+for data parallelism  \cite{chakravarty2007data}, distributed programming \cite{Murphy:2007:TDP:1793574.1793585}, reactive programming \cite{mandel2005reactiveml}, authenticated data structures \cite{Miller:2014:ADS:2535838.2535851}, databases \cite{Ohori:2011:MSM:2034773.2034815},  units of measure \cite{conf/cefp/Kennedy09} and string sanitation \cite{sanitation-psp14}.% All of these are implemented as dialects of existing languages, presumably because a strong encoding was not feasible.
 
@@
 
-This dialect-oriented state of affairs is, we argue, unsatisfying: a programmer can choose either a dialect supporting a principled approach to distributed programming, or one that builds in support for statically reasoning about units of measure, but there may not be an available dialect supporting both.
-Combining dialects is non-trivial in general, as we will discuss.
-Using different dialects separately for different components of a program is also untenable: components written in different dialects cannot always interface safely with one another (i.e.
-an FFI, item 3  above, is needed).
+This \emph{dialect-oriented} state of affairs is unsatisfying for language  designers, fragment providers and programmers alike.
+Language designers are burdened with needing to understand the complete metatheory of every fragment they add to their language to make sure it is not incompatible with existing fragments, so languages change rarely.
+This decreases the impact of many potentially useful innovations designed by fragment providers (e.g.
+many academic researchers) by making them unavailable to programmers.
+At best, programmers can choose from either a dialect supporting, for example, a principled approach to distributed programming, or one that builds in support for statically reasoning about units of measure.
+There may not be an available dialect supporting both.
+Using different dialects separately for different components of a program is untenable: components written in different dialects cannot always interface safely (an FFI, above, is needed).
 
-These problems do not arise when a fragment can be exposed as an isomorphic embedding because modern \emph{module systems} can enforce abstraction barriers that ensure that the isomorphism need only be established in the ``closed world'' of the module.
+These problems do not arise when a fragment can be expressed as an isomorphic embedding (i.e.
+as a library) because modern \emph{module systems} enforce abstraction barriers that ensure that the isomorphism need only be established in the ``closed world'' of the module.
 This is useful because it does not impose proof obligations on clients in the ``open world''.
+For example, a module defining the semantics of sets in ML can hold the representation of sets abstract, ensuring that any invariants maintained by the functions in the module (e.g.
+uniqueness) will hold even when other modules are in use.
 %Mechanisms that can help decrease the complexity of an embedding without violating abstraction barriers are thus valuable, and we will lead into our work in Sec.
 \ref{prior-work} by summarizing them.
-For situations where an embedding is not evident, however, mechanisms are needed that make specifying, implementing and reasoning about direct fragment composition similarly modular.
+
+Because embedding of this form are not always possible, as in the examples enumerated above, so mechanisms are needed that make it possible to modularly define new fragments that more directly influence the static and dynamic semantics of a language.
 Such mechanisms could ultimately be integrated directly into a language, blurring the distinction between fragments and libraries and decreasing the need for new dialects.
-Importing fragments that introduce new syntax and semantics   would be as easy as importing a new module is today.
-In the limit, the community could rely on modularly mechanized metatheory and compiler correctness results, rather than requiring heroic efforts on the part of individual research groups as the situation exists today.
+Importing fragments that introduce new syntax and semantics   would be as safe and easy as importing a new module is today.
+%In the limit, the community could rely on modularly mechanized metatheory and compiler correctness results.% rather than requiring heroic efforts from individual research groups that consider an entire language at once.
 %Recent work has shown progress on modularly introducing new concrete syntax, reviewed in Sec.
@@
 Our focus is on the problem of introducing new semantics.
+\begin{figure}[t]
+%\small
+\input{syntax-table-IL}
+\caption{Syntax of {$\mathcal{L}\{{\rightharpoonup}\,{\forall}\,{\mu}\,{\iunit}\,{\times}\,{+}\}$}, our internal language (IL).}
+\label{syntax-IL}
+%\vspace{-10px}
+\end{figure}
 
-\paragraph{Contributions} In this paper, we take substantial steps towards this goal by  constructing a minimal but powerful core calculus, @$\lambda$ (the ``actively typed'' lambda calculus).
+\vspace{-10px}
+\subsubsection{Contributions} In this paper, we take substantial steps towards this goal by  constructing a minimal but powerful core calculus, @$\lambda$ (the ``actively typed'' lambda calculus).
 %Despite its minimality, it can host a variety of practical semantic extensions like those described above, while maintaining strong metatheoretic guarantees and, crucially, providing modular reasoning properties.
@@
 \ref{overview}.
-This calculus is structured in a manner similar to the Harper-Stone semantics for Standard ML \cite{Harper00atype-theoretic}, consisting of an \emph{external language} (EL) governed by a {typed translation semantics} targeting a fixed \emph{internal language} (IL).
-Rather than building in a monolithic set of type constructors, however, the typechecking judgement is indexed by a \emph{tycon context}.
+This calculus is structured in a manner similar to the Harper-Stone semantics for Standard ML \cite{Harper00atype-theoretic}, consisting of an \emph{external language} (EL) governed by a {(bidirectionally) typed translation semantics} targeting a fixed \emph{internal language} (IL).
+Rather than building in a monolithic set of external type constructors, however, the typechecking judgement is indexed by a \emph{tycon context}.
 Each tycon defines the semantics of its associated operators via functions written in a \emph{static language} (SL).
@@
 
-We introduce @$\lambda$ by example in Sec.
-\ref{overview}, demonstrating its surprising expressive power and detailing its semantics, then examine its key metatheoretic properties in Sec.
-\ref{calculus}, beginning with \emph{type safety} and \emph{unicity of typing} and touching on  \emph{decidability of typechecking}.
-We then consider properties that relate to composition of tycon definitions: \emph{hygiene}, \emph{stability of typing} and a key modularity result, which we refer to as  \emph{conservativity}: any invariants that can be established about all values of a type under \emph{some} tycon context (i.e.
+The bulk of the paper, in Sec.
+\ref{atlam}, introduces @$\lambda$ by building up two examples, one defining labeled product types with a functional update operator,  and the other regular string types, based on a recent ``core calculus'' style specification \cite{sanitation-psp14}.
+We then examine the key metatheoretic properties of the calculus in Sec.
+\ref{metatheory}, beginning with \emph{type safety} and \emph{unicity of typing}, touching on  \emph{decidability of typechecking} and finally considering properties that relate to composition of tycon definitions: \emph{hygiene}, \emph{stability of typing} and a key modularity result, which we refer to as  \emph{conservativity}: any invariants that can be established about all values of a type under \emph{some} tycon context (i.e.
 in some  ``closed world'') are conserved in any further extended tycon context (i.e.
 in the ``open world'').
+Interestingly, the approach we take to guarantee conservativity relies on type abstraction, in this case in the internal language.
+As a result, we are able to borrow the same results that underly modular reasoning in languages like ML to reason about extensions to the semantics itself.
+Finally, we describe  related work in Sec.
+\ref{prior-work}.
 
+\section{@$\lambda$}\label{atlam}
+\begin{figure}[t]
+%\small
+\hspace{-2px}\input{syntax-table-EL}
+\caption{Syntax of the external language (EL).
+We write $x$ to range over variables and $\opname{op}$ over operator names.
+The form $\eother{\iota}$ is a technical device (and would have no concrete syntax).}\label{syntax-EL}
+\end{figure}
+
+We will begin by giving an overview of the basic organization of @$\lambda$ and introduce the main judgements in Section \ref{overview}, discuss how types are constructed and tycons are defined in Section \ref{types}, then describe how the semantics of external terms are controlled by tycons in Section \ref{external-terms}.
+
+\subsection{Overview}\label{overview}
+\vspace{-5px}
+\subsubsection{External Language}
+Programs in @$\lambda$ are written as \emph{external terms}, $e$.
+The syntax of external terms is shown in Figure \ref{syntax-EL}.
+%We will describe useful syntactic desugarings atop this syntax as we go on and review recent techniques that permit modularly introducing  desugarings like these in Sec.
+\ref{desugaring}.
+Our main focus here is on semantic extensions.
+The static and dynamic semantics of external terms are specified simultaneously as a \emph{bidirectionally typed translation semantics}.
+The two central judgements in this paper take the form:  \[\esynX{e}{\st\moutput}{\iota\moutput} ~~~~~\text{and}~~~~~ \eanaX{e}{\st}{\iota\moutput}\]
+\noindent
+
+These are pronounced ``$e$ (synthesizes / analyzes against) type $\sigma$ and has  translation $\iota$ under typing context $\Upsilon$ and tycon context $\Phi$''.
+Our specifications in this paper are intended to be algorithmic: we indicate ``outputs'' when introducing judgement forms by \emph{mode annotations}, $\moutput$; these are not part of the judgement's syntax.
+
+Bidirectional typechecking, also sometimes called \emph{local type inference} \cite{Pierce:2000:LTI:345099.345100}, is increasingly being used in modern languages (e.g.
+Scala) because it eliminates the need for type annotations in many circumstances while remaining decidable in more situations than Hindley-Milner style inference and providing what are widely perceived to be higher quality error messages, due to the locality of reasoning.
+It will also give us a clean way to reuse a fixed set of  introductory forms, discussed in Sec.
+\ref{external-terms}.
+
+%We will return to how we control the semantics of the single introductory form in the abstract syntax, $\eintro{\st}{\es}$, on the basis of the type it is being analyzed against, sidestepping issues related to the \emph{expression problem} as a result \todo{cite other paper}\cite{wadler1998expression}.
+The form $\eother{\iota}$ is a technical device used to quantify over all terms that may arise in a ``future'' tycon context, serving an analagous role to the  ``catchall'' base constants commonly found in minimal calculi, and we will return to its semantics later as well.%We will see how we leverage this bidirectionality when we discuss literal forms below.
+% The judgement is only well-defined for \emph{valid} tycon contexts, written judgementally as $\vdash \Phi$ and \emph{valid} typing contexts, $\vdash_\Phi \Upsilon$, which we also specify below.
+%  \begin{figure*}[t!]
+% \small
+% \hspace{-5px}\input{example-table}
+% %\\~\\
+% %$\begin{array}{rcl}
+% %e\cdot\conclbl{lbl}(\conclbl{lbl}_1=e_1, \cdots, \conclbl{lbl}_n=e_n)  :=  \etarg{(\conclbl{lbl}; \svar{list}n[\klbl]~\conclbl{lbl}_1~\cdots~\conclbl{lbl}_n)}{e}{e_1; \cdots; e_n}
+% %\end{array}$
+% \caption{An example external term, $e_{ex}$, in concrete syntax (left), desugared to abstract syntax (right), with static terms shown in green in examples (only).
+The notation ${\small \scolor{\desugar{...}}}$ stands for an embedding of the indicated ${\small \scolor{\conclbl{label}}}$, ${\small \scolor{\concrx{regular expression}}}$, ${\small \scolor{\concstr{string}}}$ or numeral into the SL, with derived kinds  $\klbl$, $\krx$,  $\kstr$ and $\knat$, not shown.
+The signatures of helper functions, e.g.
+${\small \scolor{\svar{nil}}}$ and ${\small \scolor{\svar{list}n}}$, are shown in Figure \ref{helper-sigs}.}
+% \label{example}
+% \end{figure*}
+
+\vspace{-10px}
+\subsubsection{Internal Language} Internal terms, $\iota$, together with internal types, $\tau$, form the \emph{typed internal language}.
+@$\lambda$ relies on a typed internal language supporting type abstraction (i.e.
+universal quantification over types).
+In this paper, we use {$\mathcal{L}\{{\rightharpoonup}\,{\forall}\,{\mu}\,{\iunit}\,{\times}\,{+}\}$}, the syntax for which is shown in Figure \ref{syntax-IL}.
+This is representative of a  standard intermediate language as might be found for a language like ML.
+The IL has a fixed semantics, so different choices of IL represent different dialects of @$\lambda$.
+
+We assume the internal statics are specified in the standard way by judgements for internal type formation {$\iDelta \vdash \tau$}, typing context formation { $\iDelta \vdash \iGamma$} and type assignment {$\iDelta~\iGamma \vdash \iota : \tau\moutput$}.
+The typing and type formation contexts obey standard structural properties (i.e.
+weakening, exchange, contraction).
+Throughout the paper, we will omit leading $\emptyset$ (used as the base case for finite mappings) and $\cdot$ (used as the base case for finite sequences) in examples (e.g.
+writing $\Gamma_\text{1} := x : \tau$ rather than $\Gamma_\text{1} := \emptyset, x : \tau$).
+
+The internal dynamics are also a standard structural operational semantics with stepping judgement {\small $\iota \mapsto \iota\moutput$} and value judgement {$\iota~\mathtt{val}$}.
+The judgement $\iota \mapsto^{*} \iota\moutput$ is the reflexive, transitive closure of the stepping judgement.
+The semantics of the IL can be found in any standard textbook covering typed lambda calculi (we closely follow \cite{pfpl}), so we assume familiarity and omit the details.
+
+% We also here define a syntax for simultaneous substitutions, of terms for variables, $\gamma$, and of types for polymorphic type variables, $\delta$, and assume standard judgements ensuring that these are valid with respect to a valid context, $\Delta \vdash \gamma : \Gamma$ and $\vdash \delta : \Delta$.
+We apply these substitutions to terms, types and typing contexts using the syntax $[\gamma]\iota$, $[\delta]\iota$, $[\delta]\tau$ and $[\delta]\Gamma$ (in some prior work, application of a substitution like this is written using a hatted form, e.g.
+$\hat\gamma(\iota)$; we intend the same semantics but use a notation more consistent with standard substitution, e.g.
+$[\iota/x]\iota'$).
+We will omit  leading $\emptyset$ and $\cdot$ in examples; in specifications, the former is used for  metatheoretic finite mappings, the latter for metatheoretic ordered lists.
+
+\vspace{-10px}
+\subsubsection{Static Language}
+\begin{figure}[t]
+\input{syntax-table-SL}
+\caption{Syntax of the static language (SL).}
+\vspace{-10px}
+\label{syntax-SL}
+\end{figure}
+The workhorse of our calculus is the \emph{static language}, which itself forms a typed lambda calculus where 
+%External terms are classified by (external) \emph{types}. 
+\emph{kinds}, $\kappa$, classify \emph{static terms}, $\sigma$.
+The syntax of the SL is given in Figure \ref{syntax-SL}.
+We note that the core of the SL is a total functional programming language based  on several standard fragments: total functions, quantification over kinds, inductive kinds (constrained by a positivity condition to prevent non-termination), and products and sums.
+These all also closely follow \cite{pfpl}, so we again omit some details.
+Only three new kinds are introduced: $\kty$, $\kity$ and $\kitm$.
+We will describe these shortly.
+We write static term variables and kind variables in bold for clarity.
+
+The kinding judgement takes the form $\sofkX{\st}{\kappa\moutput}$, where $\kDelta$ and $\kGamma$ are analagous to $\iDelta$ and $\iGamma$ and analagous well-formedness judgements $\kDelta \vdash \kappa$ and $\kDelta \vdash \kGamma$ are also defined.
+The natural number $n$ is used as a technical device in our semantics to prevent certain forms (those indexed by $n$ in the syntax above) from arising except internally in the semantics (they would have no corresponding concrete syntax); $n$ can be assumed $0$ in all user-defined terms.
+The dynamic semantics of static terms is defined as a structural operational semantics by the stepping judgement $\sstep{\st}{\argEnv}{\st\moutput}$, the value judgement $\sval{\st}{\argEnv}$ and the error judgement $\serr{\st}{\argEnv}$.
+Here, $\argEnv$ ranges over \emph{argument environments}, which we will also return to later; we omit the subscript $\argEnv$ in cases where $\argEnv = \cdot; \emptyset; \cdot$.
+We define the judgement $\smanystep{\st}{\argEnv}{\st\moutput}$ as the reflexive, transitive closure of the stepping judgement and the evaluation judgement $\seval{\st}{\argEnv}{\st'}$ iff $\smanystep{\st}{\argEnv}{\st'}$ and $\sval{\st'}{\argEnv}$.
+%${\st}{\memD}{\memG}{\aCtx}{\st\moutput}{\memD\moutput}{\memG\moutput}$, where $\memD$, $\memG$ and $\aCtx$ are also technical devices that will be described later; they too can be ignored from the perspective of user code.
+We write $\st \Downarrow \st'$ iff $\snorm{\st}{\emptyset}{\cdot}{{\rightharpoonup}; \emptyset; \emptyset}{\st'}{\emptyset}{\cdot}$.
+\emph{Static values} are normalized static terms.
+Normalization can also raise an error (to indicate a type error in an external term, or a problem in a tycon definition, as we will discuss), indicated by the judgement $\serrX{\st}$.
+We omit error propagation rules.%We will refer to the relevant rules as we proceed.
+
+\subsection{Types and Tycon Contexts}\label{types}
+
+\begin{figure}[t]
+$\begin{array}{lrcllrcl}
+\textbf{tycons} & c & ::= & \rightharpoonup ~|~ \tc\\
+\textbf{tycon contexts} & \Phi & ::= & \cdot ~|~ \Phi, \tcdef{\tc}{\psi}{\theta}\\
+\textbf{tycon structures} & \theta & ::= & \tcstruct{\st}{\omega} & \textbf{tycon sigs} & \psi & ::= & \tcsig{\kappa}{\chi}\\
+\textbf{opcon structures} & \omega & ::= & \tcstructn{\st} ~|~ \tcstructc{\omega}{op}{\st}~~~~&\textbf{opcon sigs}& \chi & ::= & \introsig{\kappa} ~|~ \opsigS{\theta}{op}{\kappa}\\
+\end{array}$
+\caption{Syntax of type constructors. Metavariable $\tc$ ranges over user-defined tycon names.}
+\label{syntax-TC}
+\end{figure}
+
+External types, or simply \emph{types}, are static values of kind $\kty$. We define the auxiliary relation $\istype{\st}{\Phi}$ for convenience:
+\begin{definition} $\istype{\st}{\Phi}$ iff $\sofkn{\emptyset}{\emptyset}{\Phi}{0}{\st}{\kty}$ and $\svalNA{\st}$.\end{definition}
+
+There are two introductory forms for  types: $\sty{c}{\st}$, where $c$ is a tycon and $\st$ is the \emph{type index}, and $\sotherty{m}{\tau}$.
+The syntax for tycons, shown in Fig.
+\ref{syntax-TC}, specifies that $c$ can either be the built-in tycon governing partial functions, $\rightharpoonup$, or a user-defined tycon name, written in small caps, $\tc$.
+There are three kinding rules governing these forms:\begin{mathpar}
+\small
+\inferrule[k-parr]{
+    \sofkX{\st}{\kprod{\kty}{\kty}}
+}{
+    \sofkX{\sty{\rightharpoonup}{\st}}{\kty}
+}
+~~~
+\inferrule[k-ty]{
+    \tcdef{\tc}{\tcsig{\ktyidx}{\chi}}{\theta} \in \Phi\\\\
+    \sofkX{\st}{\ktyidx}
+}{
+    \sofkX{\sty{\tc}{\st}}{\kty}
+}
+~~~
+\inferrule[k-otherty]{
+    \emptyset \vdash \tau
+}{
+    \sofkX{\sotherty{m}{\tau}}{\kty}
+}
+\end{mathpar}
+
+The rule \rulename{k-parr} specifies that the type index of partial function types must be a pair of types.
+We thus say that $\rightharpoonup$ has \emph{index kind} $\kty \times \kty$.
+To recover a conventional syntax, we might introduce a desugaring from $\st_1 \rightharpoonup \st_2$ to $\sty{\rightharpoonup}{(\st_1, \st_2)}$.
+%Here, we will benefit by treating it uniformly.
+
+All other tycons must be defined in the \emph{tycon context}, $\Phi$, ostensibly by the users of the language, rather than its designers.
+Each tycon specifies a \emph{tycon signature}, $\psi$, of the form $\tcsig{\ktyidx}{\chi}$, where $\ktyidx$ is the tycon's index kind (we will return to the opcon signature, $\chi$, in Sec.
+\ref{external-terms}).
+The first premise of the rule \rulename{k-ty} extracts the index kind and the second checks the type index against it.
+
+For example, we will define a type constructor $\tcvar{rstr}$ classifying \emph{regular strings}, which  are known to be in a regular language specified by a regular expression, closely following \cite{sanitation-psp14}.
+The index kind of $\tcvar{rstr}$ is $\krx$, which we assume is defined as an inductive sum kind in the standard way.
+We can write its signature, $\psi_\text{rstr}$,  and define a tycon context containing only its definition, $\Phi_\text{rstr}$,  as follows:
+\[\begin{array}{lcl}
+\psi_\text{rstr} & := & \tcsig{\krx}{\chi_\text{rstr}}\\
+\Phi_\text{rstr} & := & \tcdef{\tcvar{rstr}}{\psi_\text{rstr}}{\theta_\text{rstr}}
+\end{array}\]
+We then have that, e.g., $\sofkz{\emptyset}{\emptyset}{\Phi_\text{rstr}}{\stx{title}}{\kty}$ and $\sofkz{\emptyset}{\emptyset}{\Phi_\text{rstr}}{\stx{conf}}{\kty}$, where 
+\[
+\begin{array}{lcl}
+\stx{title} & := & \sty{\tcvar{rstr}}{\concrx{.+}}\\
+\stx{conf}  & := & \sty{\tcvar{rstr}}{\concrx{[A-Z]+ \digit\digit\digit\digit}}
+\end{array}\]
+The type index is here written using standard concrete syntax for regular expressions for concision.
+In previous work, we have shown how to define type-specific (here, kind-specific) syntax like this composably in libraries \cite{TSLs}.
+
+The second example we will define is the tycon $\tcvar{lprod}$, which will define a variant of labeled product type (which are like record types, but maintain a row  ordering).
+The index kind of $\tcvar{lprod}$ is $\klist{\klbl \times \kty}$, assuming lists are defined in the usual way, and $\klbl$ classifies static representations of syntactic labels.
+We can write the tycon signature of $\tcvar{lprod}$, $\psi_\text{lprod}$,  and the tycon context containing only its definition, $\Phi_\text{lprod}$, as follows:
+\[\begin{array}{lcl}
+\psi_\text{lprod} & := & \tcsig{\klist{\klbl \times \kty}}{\chi_\text{lprod}}\\
+\Phi_\text{lprod}  & := & \tcdef{\tcvar{lprod}}{\psi_\text{lprod}}{\theta_\text{lprod}}
+\end{array}\]
+
+In a tycon context containing both these tycon definitions, $\Phi_\text{rstr}\Phi_\text{lprod}$, we can derive that $\sofkz{\emptyset}{\emptyset}{\Phi_\text{rstr}\Phi_\text{lprod}}{\stx{paper}}{\kty}$ where: \[\small\stx{paper} := \sty{\tcvar{lprod}}{\{\conclbl{title} : \stx{title}, \conclbl{conf} : \stx{conf}\}}\]
+
+We again use kind-specific syntax, here for $\klbl$ and $\klist{\klbl \times \kty}$,  for concision, and to again demonstrate how standard syntax for types can be recovered composably despite the uniform abstract syntax for types we use in our specification of the language.
+
+The remaining introductory form for types, $\sotherty{m}{\tau}$, is a technical device that will be used to quantify over all types other than those in a given tycon context.
+As there may be arbitrarily many of these, the term is indexed by a natural number, $m$.
+It is also indexed by a closed internal type, $\tau$, which serves as its translation, discussed below.
+
+The operational semantics for these forms is straightforward: the type index is recursively evaluated to a value and errors are propagated:
+\begin{mathpar}
+\small
+\inferrule[s-ty-step]{
+    \sstep{\st}{\argEnv}{\st'}
+}{
+    \sstep{\sty{c}{\st}}{\argEnv}{\sty{c}{\st'}}
+}
+
+\inferrule[s-ty-err]{
+    \serr{\st}{\argEnv}
+}{
+    \serr{\sty{c}{\st}}{\argEnv}
+}
+
+\inferrule[s-ty-v]{
+    \sval{\st}{\argEnv}
+}{
+    \sval{\sty{c}{\st}}{\argEnv}
+}
+
+\inferrule[s-otherty-v]{ }{
+    \sval{\sotherty{m}{\tau}}{\argEnv}
+}
+\end{mathpar}
+
+\subsubsection{Type Case Analysis}
+A type $\st$ can be case analyzed against a known tycon $c$ using $\stycase{c}{\st}{\sx}{\st_1}{\st_2}$.
+If the value of $\st$ is constructed by $c$, its type index is bound to $\sx$ and the branch $\st_1$ is taken.
+For totality, a default branch, $\st_2$, must also be provided.
+\begin{mathpar}
+\small
+\inferrule[k-tycase-parr]{
+    \sofkX{\st}{\kty}\\\\
+    \sofk{\kDelta}{\kGamma, \sx :: \kty \times \kty}{\Phi}{\st_1}{\kappa}\\\\
+    \sofkX{\st_2}{\kappa}
+}{
+    \sofkX{\stycase{\rightharpoonup}{\st}{\sx}{\st_1}{\st_2}}{\kappa}
+}
+
+\inferrule[k-tycase]{
+    \sofkX{\st}{\kty}\\\\
+    \tcdef{\tc}{\tcsig{\ktyidx}{\theta}}{\opdefs} \in \Phi\\\\
+    \sofk{\kDelta}{\kGamma, \sx :: \ktyidx}{\Phi}{\st_1}{\kappa}\\
+    \sofkX{\st_2}{\kappa}
+}{
+    \sofkX{\stycase{\tc}{\st}{\sx}{\st_1}{\st_2}}{\kappa}
+}
+\end{mathpar}
+The operational semantics are straightforward: 
+\begin{mathpar}
+\small
+\inferrule[s-tycase-step]{
+    \sstep{\st}{\argEnv}{\st'}
+}{
+    \sstep{\stycase{c}{\st}{\sx}{\st_1}{\st_2}}{\argEnv}{\stycase{c}{\st'}{x}{\st_1}{\st_2}}
+}
+
+\inferrule[s-tycase-err]{
+    \serr{\st}{\argEnv}
+}{
+    \serr{\stycase{c}{\st}{\sx}{\st_1}{\st_2}}{\argEnv}
+}
+\end{mathpar}
+\begin{mathpar}\small
+\inferrule[s-tycase-match]{
+    \sval{\sty{c}{\st}}{\argEnv}
+}{
+    \sstep{\stycase{c}{\sty{c}{\st}}{\sx}{\st_1}{\st_2}}{\argEnv}{[\st/x]\st_1}
+}
+
+\inferrule[s-tycase-fail]{
+    \st \neq \sty{c}{\st'}
+}{
+    \sstep{\stycase{c}{\st}{\sx}{\st_1}{\st_2}}{\argEnv}{\st_2}
+}
+\end{mathpar}
+% \begin{mathpar}\small
+% \inferrule[s-tycase-fail-2]{ }{
+%     \sstep{\stycase{c}{\sotherty{m}{\tau}}{x}{\st_1}{\st_2}}{\argEnv}{\st_2}
+% }
+% \end{mathpar}
+
+Put another way, types can be thought of as arising from a distinguished ``open datatype'' defined by the tycon context.
+We will see how case analysis can be useful for defining the semantics of operators associated with tycons in Sec.
+\ref{external-terms}.
+
+\subsubsection{Tycon Context Formation}
+The tycon context formation judgement, $\vdash \Phi$, requires that tycon names are unique (we assume some extralinguistic mechanism is used for avoiding naming conflicts) and checks that each tycon structure, $\theta$, is valid against its tycon signature, $\psi$.
+We consider the premises of (tcc-ext) below:
+\begin{mathpar}
+\small
+\inferrule[tcc-emp]{ }{
+    \vdash \cdot
+}
+~~~~
+\inferrule[tcc-ext]{
+    \vdash \Phi\\
+    \tc \notin \text{dom}(\Phi)\\
+    \theta = (\tcstruct{\stx{schema}}{\omega})\\
+    \psi = (\tcsig{\ktyidx}{\chi})\\\\
+    \keq{\emptyset}{\ktyidx}\\
+    \sofkz{\emptyset}{\emptyset}{\Phi}{\stx{schema}}{\ktyidx \rightarrow \kity}\\
+    \vdash_{\Phi, \tcdef{\tc}{\psi}{\theta}} \omega \sim \psi
+}{
+    \vdash \Phi, \tcdef{\tc}{\psi}{\theta}
+}
+\end{mathpar}
+
+\subsubsection{Type Equivalence} 
+To simplify the handling of type equivalence, type index kinds must be \emph{equality kinds}: those for which semantic equivalence implies syntactic equality of values.
+We define these by the judgement $\keq{\kDelta}{\kappa}$, appearing as a premise in (tcc-ext).
+Equality kinds are similar to equality types as found in Standard ML. The main implication of this choice is that type indices cannot contain static functions.
+\begin{mathpar}
+\small
+\inferrule[keq-k]{\kvar \in \kDelta}{\keq{\kDelta}{\kvar}}
+
+\inferrule[keq-ind]{\keq{\kDelta, \kvar}{\kappa}}{
+  \keq{\kDelta}{\kmu{\kvar}{\kappa}}}
+
+\inferrule[keq-unit]{ }{\keq{\kDelta}{\kunit}}
+\\
+\inferrule[keq-prod]{
+  \keq{\kDelta}{\kappa_1}~~~~
+  \keq{\kDelta}{\kappa_2}
+}{
+  \keq{\kDelta}{\kprod{\kappa_1}{\kappa_2}}
+}
+
+\inferrule[keq-sum]{
+  \keq{\kDelta}{\kappa_1}~~~~
+  \keq{\kDelta}{\kappa_2}
+}{
+  \keq{\kDelta}{\ksum{\kappa_1}{\kappa_2}}
+}
+
+\inferrule[keq-ty]{ }{\keq{\kDelta}{\kty}}
+\end{mathpar}
+ %We will discuss strategies for fragments that benefit from more sophisticated equivalences in Section \ref{examples}.\todo{do this}
+
+\subsubsection{Type Translations}
+Our style of specifying the meaning of external terms directly in terms of a  translation to the IL is similar to the Harper-Stone \emph{typed elaboration semantics} for Standard ML \cite{Harper00atype-theoretic}.
+There, however, external and internal terms were governed by a common type system.
+In @$\lambda$, each external type maps onto an internal type, called its \emph{translation}, as specified by the judgement $\vdash_\Phi \st \leadsto \tau$, defined below.
+This style may thus also be compared to specifications for the first stage of a type-directed compiler, e.g.
+the TIL compiler for Standard ML \cite{tarditi+:til-OLD}, here lifted ``one level up'' into the semantics of the language itself.
+
+Each tycon determines the translations of the types it constructs as a function of each type's index by specifying a \emph{translation schema} in the {tycon structure}.
+For a tycon with index kind $\ktyidx$, the translation schema has kind $\ktyidx \rightarrow \kity$, checked by (tcc-ext).
+
+The kind $\kity$ has a single introductory form, $\sqity{\qity}$, where $\qity$ is a \emph{translational internal type}.
+Each form in the syntax for internal types, $\tau$,  corresponds to a form in the syntax of translational internal types, $\qity$.
+For example, our translation schema for $\tcvar{rstr}$  simply chooses to ignore the type index and represent all regular strings internally as strings, of internal type abbreviated $\keyw{str}$.
+We abbreviate the corresponding translational internal type $\hat{\keyw{str}}$ and define the translation schema as follows:
+\[
+\begin{array}{lcl}
+\theta_\text{rstr} := \tcstruct{\slam{\krx}{\svar{tyidx}}{\sqity{\hat{\keyw{str}}}}}{\omega_\text{rstr}}
+\end{array}
+\]
+
+The kinding rules for these shared forms simply proceed recursively, e.g.:
+\begin{mathpar}
+\small
+\inferrule[k-ity-lam]{
+    \sofkX{\sqity{\qity_1}}{\kity}\\
+    \sofkX{\sqity{\qity_2}}{\kity}
+}{
+    \sofkX{\sqity{\qity_1 \times \qity_2}}{\kity}
+}
+
+\inferrule[k-ity-alpha]{ }{\sofkX{\sqity{\alpha}}{\kity}}
+\end{mathpar}
+The operational semantics for shared forms are also straightforwardly recursive:
+\begin{mathpar}
+\small
+\inferrule[s-ity-lam-step-1]{
+    \sstep{\sqity{\qity_1}}{\argEnv}{\sqity{\qity_1'}}%\sstep{\sqity{\qity_1}}{\argEnv}{\sstep{\sqity{\qity_1'}}}
+}{
+    \sstep{\sqity{\qity_1\times\qity_2}}{\argEnv}{\sqity{\qity_1' \times \qity_2}}
+}
+
+\inferrule[s-ity-lam-step-2]{
+    \sval{\sqity{\qity_1}}{\argEnv}\\
+    \sstep{\sqity{\qity_2}}{\argEnv}{\sqity{\qity_2'}}%\sstep{\sqity{\qity_1}}{\argEnv}{\sstep{\sqity{\qity_1'}}}
+}{
+    \sstep{\sqity{\qity_1\times\qity_2}}{\argEnv}{\sqity{\qity_1 \times \qity_2'}}
+}
+
+\inferrule[s-ity-lam-err-1]{
+    \serr{\sqity{\qity_1}}{\argEnv}
+}{
+    \serr{\sqity{\qity_1\times\qity_2}}{\argEnv}
+}
+
+\inferrule[s-ity-lam-err-2]{
+    \serr{\sqity{\qity_2}}{\argEnv}
+}{
+    \serr{\sqity{\qity_1\times\qity_2}}{\argEnv}
+}
+
+\inferrule[s-ity-lam-v]{
+    \sval{\sqity{\qity_1}}{\argEnv}\\
+    \sval{\sqity{\qity_2}}{\argEnv}
+}{
+    \sval{\sqity{\qity_1\times\qity_2}}{\argEnv}
+}
+
+\inferrule[s-ity-alpha-v]{ }{
+    \sval{\sqity{\alpha}}{\argEnv}
+}
+\end{mathpar}
+
+The syntax for translational internal types additionally includes an ``unquote'' form,  $\qtuq{\st}$, so that they can be constructed compositionally, as well as a form, $\srep{\st}$, that refers to the translation of type $\st$.
+For example, our translation schema for $\tcvar{lprod}$ translates labeled product types to nested binary product types computed by folding over the type index and referring to the translations of the types  therein.
+We assume a standard list recursor, $\small\svar{listfold} :: \kforall{\kalpha_1}{\kforall{\kalpha_2}{\klist{\kalpha_1}\rightarrow\kalpha_2\rightarrow(\kalpha_1\rightarrow\kalpha_2\rightarrow\kalpha_2)\rightarrow\kalpha_2}}$:
+\[\small
+\begin{array}{lcl}
+\theta_\text{lprod} & := & \tcstruct{\slam{\klist{\klbl \times \kty}}{\svar{tyidx}}{\\
+&&\quad\skap{\kity}{\skap{\klbl\times\kty}{\svar{listfold}}}~\svar{tyidx}~\sqity{\iunit}~\\&&
+  \quad\quad (\lambda \svar{h}{:}\klbl \times \kty.\lambda \svar{r}{:}\kity.\keyw{letpair}~(\svar{hlbl},\svar{hty})=\svar{h}~\keyw{in}~\\&&
+  \quad\quad\quad \sqity{\srep{\svar{hty}}\times\qtuq{\svar{r}}}}\\&&}{\omega_\text{rstr}}
+\end{array}\]
+For example, evaluating this translation schema with the index of $\stx{paper}$ produces $\stx{paper/trans} := \sqity{\qity_\text{paper/abstrans}}$ where: $\small\qity_\text{paper/abstrans} := \srep{\stx{title}}\times(\srep{\stx{conf}}\times \iunit)$.
+We do not include logic to optimize away the trailing unit type for simplicity.%For simplicity, and to make the point that the choice of representation has only tycon-local implications, our translation schema does not optimize away
+
+The kinding rules for the unquote and translation forms are straightforward:
+\begin{mathpar}
+\inferrule[k-ity-unquote]{
+    \sofkX{\st}{\kity}   
+}{
+    \sofkX{\sqity{\qtuq{\st}}}{\kity}
+}
+
+\inferrule[k-ity-trans]{
+    \sofkX{\st}{\kty}
+}{
+    \sofkX{\sqity{\srep{\st}}}{\kity}
+}
+\end{mathpar}
+The unquote form is eliminated during evaluation:
+\begin{mathpar}
+\small
+\inferrule[s-ity-unquote-step]{
+    \sstep{\st}{\argEnv}{\st'}
+}{
+    \sstep{\sqity{\qtuq{\st}}}{\argEnv}{\sqity{\qtuq{\st'}}}
+}
+
+\inferrule[s-ity-unquote-err]{
+    \serr{\st}{\argEnv}
+}{
+    \serr{\sqity{\qtuq{\st}}}{\argEnv}
+}
+
+\inferrule[s-ity-unquote-elim]{
+    \sval{\sqity{\qity}}{\argEnv}
+}{
+    \sstep{\sqity{\qtuq{\sqity{\qity}}}}{\argEnv}{\sqity{\qity}}
+}
+\end{mathpar}
+References to the translation of a type, however, are retained in values of kind $\kity$:
+\begin{mathpar}
+\small
+\inferrule[s-ity-trans-step]{
+    \sstep{\st}{\argEnv}{\st'}
+}{
+    \sstep{\sqity{\srep{\st}}}{\argEnv}{\sqity{\srep{\st'}}}
+}
+
+\inferrule[s-ity-trans-err]{
+    \serr{\st}{\argEnv}
+}{
+    \serr{\sqity{\srep{\st}}}{\argEnv}
+}
+
+\inferrule[s-ity-trans-val]{
+    \sval{\st}{\argEnv}
+}{
+    \sval{\sqity{\srep{\st}}}{\argEnv}
+}
+\end{mathpar}
+
+This allows us to selectively hold these translations abstract and will be the key to our main results.
+The selective type abstraction judgement $\tdeabs{\Phi}{\tc}{\qity}{\memD}{\ity\moutput}{\memD\moutput}$ relates a normalized translational internal type $\qity$ to an internal type $\tau$, called an \emph{abstract type translation} because all references to translations of types constructed by a user-defined tycon other than those constructed by a ``delegated tycon'', $\tc$, are replaced by universally quantified type variables, $\alpha$.
+Each unique type has a unique type variable associated with it, tracked by the \emph{type translation store}, $\memD$, which maps from types, $\st$, to their (concrete) translation, $\tau$, and the type variable, $\alpha$, which appears in its place in the abstract type translation: $\memD ::= \emptyset ~|~ \memD, \st \leftrightarrow \ity/\alpha$.
+
+For example, $\small\tdeabs{\Phi_\text{rstr}\Phi_\text{lprod}}{\tcvar{lprod}}{\qity_\text{paper/trans}}{\emptyset}{\tau_\text{paper/abstrans}}{\memD_\text{paper/abstrans}}$
+where: \[\begin{array}{lcl}\small
+\tau_\text{paper/abstrans} & := & \alpha_1 \times (\alpha_2 \times \iunit)\\
+\memD_\text{paper/abstrans} & := & \stx{title} \leftrightarrow \keyw{str}/\alpha_1, \stx{conf} \leftrightarrow \keyw{str}/\alpha_2
+\end{array}\]
+
+Each type translation store induces a \emph{type substitution}, $\delta$, and a corresponding internal type formation context, $\Delta$, according to the judgement $\memD \leadsto \delta : \Delta$.
+Type substitutions are a standard type theoretic construct, $\delta ::= \emptyset ~|~ \delta, \tau/\alpha$.
+
+For example, $\memD_\text{paper/abstrans} \leadsto \delta_\text{paper/abstrans} : \Delta_\text{paper/abstrans}$ where:
+\[\begin{array}{lcl}
+\delta_\text{paper/abstrans} & := & \keyw{str}/\alpha_1, \keyw{str}/\alpha_2\\
+\Delta_\text{paper/abstrans} & := & \alpha_1, \alpha_2\end{array}\]
+
+This judgement is defined by the following straightforward rules:
+\begin{mathpar}\small
+\inferrule[tstore-emp]{ }{\emptyset \leadsto \emptyset : \emptyset}
+
+\inferrule[tstore-ext]{\memD \leadsto \delta : \Delta}{(\memD, \st \leftrightarrow \tau/\alpha) \leadsto (\delta, \tau/\alpha) : (\Delta, \alpha)}
+\end{mathpar}
+
+We can apply type substitutions to internal types, terms and typing contexts, written $[\delta]\ity$, $[\delta]\iota$ and $[\delta]\Gamma$, respectively.
+For example, $[\delta_\text{paper/abstrans}]\tau_\text{paper/abstrans}$ is: \[
+\begin{array}{lcl}
+\tau_\text{paper} := \keyw{str} \times (\keyw{str} \times \iunit)\end{array}\]
+
+This process of selectively holding the translations of all types constructed by any tycon other than the ``delegated tycon'' abstract will be critical for conservativity to hold below.
+This can be thought of as analagous to the process in ML by which the true identity of an abstract type in a module is held abstract outside the module until after typechecking is complete.
+Here, the translation of the type is being held abstract, rather than the type itself, and type translations are computed from indices, rather than declared ``literally''.
+
+The selective type abstraction judgement is defined by recursing generically over sub-terms until terms of the form $\srep{\st}$ are encountered. For example, 
+\begin{mathpar}
+\small
+\inferrule[abs-iparr]{
+    \tdeabs{\Phi}{\tc}{\qity_1}{\memD}{\ity_1}{\memD'}\\
+    \tdeabs{\Phi}{\tc}{\qity_2}{\memD'}{\ity_2}{\memD''}
+}{
+    \tdeabs{\Phi}{\tc}{\qity_1 \times \qity_2}{\memD}{\ity_1 \times \ity_2}{\memD''}
+}
+
+\inferrule[abs-alpha]{ }{
+    \tdeabs{\Phi}{\tc}{\alpha}{\memD}{\alpha}{\memD}
+}
+\end{mathpar}
+
+The translation of partial function types is not held abstract, so that lambdas can be used as the sole binding construct in the language:
+\begin{mathpar}
+\small
+\inferrule[abs-parr]{
+    \tdeabs{\Phi}{\tc}{\srep{\st_1}}{\memD}{\ity_1}{\memD'}\\
+    \tdeabs{\Phi}{\tc}{\srep{\st_2}}{\memD'}{\ity_2}{\memD''}
+}{
+    \tdeabs{\Phi}{\tc}{\srep{\sty{\rightharpoonup}{(\st_1, \st_2)}}}{\memD}{\tau_1 \rightharpoonup \tau_2}{\memD''}
+}
+\end{mathpar}
+
+The translation of user-defined types constructed by the delegated tycon is determined by calling the translation schema and checking that the type translation it generates refers only to type variables generated from $\memD'$.
+\begin{mathpar}
+\small
+\inferrule[abs-tc-local]{
+    \tcdef{\tc}{\psi}{\tcstruct{\stx{schema}}{\omega}} \in \Phi\\\\
+    \sevalNA{\stx{schema}(\sttyidx)}{\sqity{\qity}}\\
+    \tdeabs{\Phi}{\tc}{\qity}{\memD}{\ity}{\memD'}\\
+    \memD' \leadsto \delta : \Delta\\
+    \Delta \vdash \tau
+}{
+    \tdeabs{\Phi}{\tc}{\srep{\sty{\tc}{\sttyidx}}}{\memD}{\tau}{\memD'}
+}
+\end{mathpar}
+
+The translation of a user-defined type constructed by any tycon other than the delegated tycon is added to the store, or retrieved from it if already there (using a store ensures that equal types map to equal abstract type variables, even if their translation is requested multiple times).
+The translation of $\sotherty{m}{\tau}$ is simply $\tau$.
+It is always held abstract, because it is designed to behave indistinguishably from a type from the ``future'', as we will discuss.
+\begin{mathpar}
+\small
+\inferrule[abs-tc-foreign-new]{
+    \tc \neq \tc'\\
+    \sty{\tc'}{\sttyidx} \notin \text{dom}(\memD)\\
+    \tcdef{\tc'}{\psi}{\tcstruct{\stx{schema}}{\omega}} \in \Phi\\\\
+    \sevalNA{\stx{schema}~\sttyidx}{\sqity{\qity}}\\
+    \tdeabs{\Phi}{\tc}{\qity}{\memD}{\ity}{\memD'}\\
+    \memD' \leadsto \delta : \Delta\\
+    \Delta \vdash \tau\\
+    (\alpha~\text{fresh})
+}{
+    \tdeabs{\Phi}{\tc}{\srep{\sty{\tc'}{\sttyidx}}}{\memD}{\alpha}{\memD', \sty{\tc'}{\sttyidx} \leftrightarrow \tau/\alpha}
+}
+
+\inferrule[abs-tc-other-new]{
+    \sotherty{m}{\tau} \notin \text{dom}(\memD)\\
+    (\alpha~\text{fresh})
+}{
+    \tdeabs{\Phi}{\tc}{\srep{\sotherty{m}{\tau}}}{\memD}{\alpha}{\memD, \sotherty{m}{\tau}\leftrightarrow\tau/\alpha}
+}
+
+\inferrule[abs-tc-stored]{
+    \st \leftrightarrow \tau/\alpha \in \memD
+}{
+    \tdeabs{\Phi}{\tc}{\srep{\st}}{\memD}{\alpha}{\memD}
+}
+\end{mathpar}
+
+Finally, we can give the rule defining the concrete translation judgement, $\vdash_\Phi \st \leadsto \tau$, mentioned at the beginning of this subsection.
+We simply determine any selectively abstract translation, then apply the substitution induced by the store:
+\begin{mathpar}
+\inferrule[conc-ty-trans]{
+    \tdeabs{\Phi}{\tc}{\srep{\st}}{\emptyset}{\tau}{\memD}\\
+    \memD \leadsto \delta : \Delta
+}{
+    \vdash_\Phi \st \leadsto [\delta]\tau
+}
+\end{mathpar}
+
+\subsubsection{Typing Contexts}
+The translation judgement for external typing contexts $\vdash_\Phi \Upsilon \leadsto \Gamma$ simply checks that $\Upsilon$ actually maps variables to types, then generates the corresponding internal typing context $\Gamma$.
+We follow standard practice in assuming implicitly that terms are congruent up to $\alpha$-equivalence so all variables in typing contexts can be assumed unique implicitly.
+We also have that external typing contexts obey the standard structural congruences: weakening, exchange and contraction.
+\begin{mathpar}\small
+\inferrule[etctx-emp]{ }{\vdash_\Phi \emptyset \leadsto \emptyset}
+
+\inferrule[etctx-ext]{\vdash_\Phi \Upsilon \leadsto \Gamma\\\istype{\st}{\Phi}\\\vdash_\Phi \st \leadsto \tau}{\vdash_\Phi \Upsilon, x \Rightarrow \st \leadsto \Gamma, x : \tau}
+\end{mathpar}
+
+\begin{figure*}[t]
+\fbox{$\esynX{e}{\st\moutput}{\iota\moutput}$}
+~\fbox{$\eanaX{e}{\st}{\iota\moutput}$}
+\begin{mathpar}
+\inferrule[subsume]{
+  \esynX{e}{\st}{\iota}
+}{
+  \eanaX{e}{\st}{\iota}
+}
+
+\inferrule[ascribe]{
+  \sofkz{\emptyset}{\emptyset}{\Phi}{\st}{\kty}\\
+  \snormT{\st}{\st'}\\\\
+  \eanaX{e}{\st'}{\iota}
+}{
+  \esynX{\easc{e}{\st}}{\st'}{\iota}
+}
+
+% \inferrule[syn-slet]{
+%   \tccok{\Phi}{\kXi}\\
+%   \sofkz{\emptyset}{\emptyset}{\kXi}{\st}{\kappa}\\\\
+%   \snormT{\st}{\st'}\\
+%   \esynX{[\st'/\sx]e}{\stx{ty}}{\iota}
+% }{
+%   \esynX{\eslet{\st}{\sx}{e}}{\stx{ty}}{\iota}
+% }
+
+% \inferrule[ana-slet]{
+%   \tccok{\Phi}{\kXi}\\
+%   \sofkz{\emptyset}{\emptyset}{\kXi}{\st}{\kappa}\\\\
+%   \snormT{\st}{\st'}\\
+%   \eanaX{[\st'/\sx]e}{\stx{ty}}{\iota}
+% }{
+%   \eanaX{\eslet{\st}{\sx}{e}}{\stx{ty}}{\iota}
+% }
+%
+\inferrule[syn-var]{
+  x \Rightarrow \st \in \Upsilon
+}{
+  \esynX{x}{\st}{x}
+}
+
+% \inferrule[syn-let]{
+%   \esynX{e_1}{\st_1}{\iota_1}\\\sfinalrepX{\st_1}{\tau_1}\\\\
+%   \esyn{\Upsilon, x \Rightarrow \st_1}{\Phi}{e_2}{\st_2}{\iota_2}\\
+% }{
+%   \esynX{\elet{e_1}{x}{e_2}}{\st_2}{\iap{\ilam{\tau_1}{x}{\iota_2}}{\iota_1}}
+% }
+
+% \inferrule[ana-let]{
+%   \esynX{e_1}{\st_1}{\iota_1}\\\sfinalrepX{\st_1}{\tau_1}\\\\
+%   \eana{\Upsilon, x \Rightarrow \st_1}{\Phi}{e_2}{\st_2}{\iota_2}\\
+% }{\eanaX{\elet{e_1}{x}{e_2}}{\st_2}{\iap{\ilam{\tau_1}{x}{\iota_2}}{\iota_1}}}
+%
+% \inferrule[syn-lam]{
+%   \tccok{\Phi}{\kXi}\\
+%   \sofkz{\emptyset}{\emptyset}{\kXi}{\st_1}{\kty}\\
+%   \snormT{\st_1}{\st_1'}\\
+%   \sfinalrepX{\st_1'}{\tau_1}\\\\
+%   \esyn{\Upsilon, x \Rightarrow \st_1'}{\Phi}{e}{\st_2}{\iota}
+% }{
+%   \esynX{\elam{\st_1}{x}{e}}{\sparr{(\st_1',\st_2)}}{\ilam{\tau_1}{x}{\iota}}
+% }
+\inferrule[ana-lam]{
+    \eana{\Upsilon, x \Rightarrow \st_1}{\Phi}{e}{\st_2}{\iota}\\\\
+    \sfinalrepX{\st_1}{\tau_1}
+}{
+    \eanaX{\eanalam{x}{e}}{\sparr{(\st_1, \st_2)}}{\ilam{\tau_1}{x}{\iota}}
+}
+
+\inferrule[syn-ap]{
+  \esynX{e_1}{\sparr{(\st_1, \st_2)}}{\iota_1}\\\\
+  \eanaX{e_2}{\st_2}{\iota_2}
+}{
+  \esynX{\eap{e_1}{e_2}}{\st_2}{\iap{\iota_1}{\iota_2}}
+}
+
+\inferrule[ana-fix]{
+  \eana{\Upsilon, x \Rightarrow \st}{\Phi}{e}{\st}{\iota}\\\\
+  \sfinalrepX{\st}{\tau}
+}{
+  \eanaX{\efix{x}{e}}{\st}{\ifix{\tau}{x}{\iota}}
+}
+
+\inferrule[ana-lit]{
+  \tcdef{\tc}{\tcsig{\ktyidx}{\theta}}{\omega} \in \Phi\\\\
+  \introsig{\klitidx} \in \theta\\
+  \sofkz{\emptyset}{\emptyset}{\Phi}{\stmidx}{\klitidx}\\\\
+  \keyw{ana intro}={\stx{def}} \in \omega\\
+  \keyw{args}(\es)=\stx{args}\\\\
+  \stx{def}~\sttyidx~\stmidx~\stx{args} \Downarrow_{\es;\Upsilon;\Phi} \sqitm{\qitm}\\\\
+  %\snorm{\stx{def}~\sttyidx~\stmidx~\stx{args}}{\emptyset}{\memG_0}{{\rightharpoonup}, \tc; \Upsilon; \Phi}{\sqitm{\itm_\text{abs}}}{\memD}{\memG}\\\\
+  %\snorm{\srep{\sty{\tc}{\sttyidx}}}{\memD}{\cdot}{{\rightharpoonup}, \tc;\emptyset;\Phi}{\sqity{\ity_\text{abs}}}{\memD'}{\cdot}\\\\
+  \qitm ... \delta:\Delta; \gamma:\Gamma; \iota\\
+  \srep{\sty{\tc}{\sttyidx}} ... \delta':\Delta'; \tau\\\\
+  \iDelta\iDelta'~\iGamma \vdash \iota : \tau
+}{
+  \eanaX{\eintro{\stmidx}{\es}}{\sty{\tc}{\sttyidx}}{[{\delta}][\delta'][{\gamma}]\iota}
+}
+
+\inferrule[syn-targ]{
+  \esynX{e_\text{targ}}{\sty{\tc}{\sttyidx}}{\iota_\text{targ}}\\\\
+  \tcdef{\tc}{\tcsig{\ktyidx}{\Omega}}{\omega} \in \Phi\\\\
+  \opsig{\opname{op}}{\ktargidx} \in \Omega\\
+  \tccok{\Phi}{\kXi}\\
+  \sofkz{\emptyset}{\emptyset}{\kXi}{\stmidx}{\ktargidx}\\\\
+  \synop{\opname{op}}{\stx{def}} \in \omega\\
+  \mkargs{e_\text{targ}; \es}{\stx{args}}{\memG_0}{n}\\\\
+  %\snorm{\srep{\sty{\tc}{\sttyidx}}}{\emptyset}{\cdot}{{\rightharpoonup}, \tc;\emptyset;\Phi}{\sqity{\ity_\text{targ}}}{\memD_0}{\cdot}\\\\
+  \snorm{\stx{def}~\sttyidx~(\svar{tr}~\ssyn{0})~\stmidx~(\svar{atl}~\stx{args})}{\emptyset}{\memG_0}{{\rightharpoonup},\tc; \Upsilon; \Phi}{(\st, \sqitm{\itm_\text{abs}})}{\memD}{\memG}\\\\
+  \snorm{\srep{\st}}{\memD}{\cdot}{{\rightharpoonup},\tc;\emptyset;\Phi}{\sqity{\ity_\text{abs}}}{\memD'}{\cdot}\\\\
+  \vdash_\Phi \memD' \leadsto {\delta} : \iDelta\\
+  \memG \leadsto {\gamma} : \iGamma\\
+  \iDelta~\iGamma \vdash \iota_\text{abs} : \tau_\text{abs}
+}{
+  \esynX{\etarg{\opname{op}}{\sttmidx}{e_\text{targ}}{\es}}{\st}{[\delta][\gamma]\iota_\text{abs}}
+}
+
+\inferrule[ana-other]{
+  \sfinalrep{\Phi}{\srep{\Upsilon}}{\Gamma \parallel \memD}\\
+  \memD \leadsto \delta; \Delta\\
+  \Delta~\Gamma \vdash \iota : \tau
+}{
+  \eanaX{\eother{\iota}}{\sotherty{n}{\tau}}{[\delta]\iota}
+}
+\end{mathpar}
+\caption{Typing}
+\label{statics-EL}
+\end{figure*}
+\subsection{Typing and Translation of External Terms}\label{external-terms}
+Having established how types are constructed, and how they determine translations to internal types, we can finally give the typing and translation rules for external terms.
+
+\subsubsection{Variables and Functions}
+Variables and functions behave in the standard manner.
+We use Plotkin's fixpoint operator to support recursive (i.e.
+partial) functions (cf.
+\cite{pfpl}), and define both only in analytic positions for simplicity:
+\begin{mathpar}
+\small
+\inferrule[syn-var]{
+  x \Rightarrow \st \in \Upsilon
+}{
+  \esynX{x}{\st}{x}
+}
+
+% \inferrule[syn-let]{
+%   \esynX{e_1}{\st_1}{\iota_1}\\\sfinalrepX{\st_1}{\tau_1}\\\\
+%   \esyn{\Upsilon, x \Rightarrow \st_1}{\Phi}{e_2}{\st_2}{\iota_2}\\
+% }{
+%   \esynX{\elet{e_1}{x}{e_2}}{\st_2}{\iap{\ilam{\tau_1}{x}{\iota_2}}{\iota_1}}
+% }
+
+% \inferrule[ana-let]{
+%   \esynX{e_1}{\st_1}{\iota_1}\\\sfinalrepX{\st_1}{\tau_1}\\\\
+%   \eana{\Upsilon, x \Rightarrow \st_1}{\Phi}{e_2}{\st_2}{\iota_2}\\
+% }{\eanaX{\elet{e_1}{x}{e_2}}{\st_2}{\iap{\ilam{\tau_1}{x}{\iota_2}}{\iota_1}}}
+%
+% \inferrule[syn-lam]{
+%   \tccok{\Phi}{\kXi}\\
+%   \sofkz{\emptyset}{\emptyset}{\kXi}{\st_1}{\kty}\\
+%   \snormT{\st_1}{\st_1'}\\
+%   \sfinalrepX{\st_1'}{\tau_1}\\\\
+%   \esyn{\Upsilon, x \Rightarrow \st_1'}{\Phi}{e}{\st_2}{\iota}
+% }{
+%   \esynX{\elam{\st_1}{x}{e}}{\sparr{(\st_1',\st_2)}}{\ilam{\tau_1}{x}{\iota}}
+% }
+\inferrule[ana-lam]{
+    \eana{\Upsilon, x \Rightarrow \st_1}{\Phi}{e}{\st_2}{\iota}\\
+    \sfinalrepX{\st_1}{\tau_1}
+}{
+    \eanaX{\eanalam{x}{e}}{\sty{\rightharpoonup}{(\st_1, \st_2)}}{\ilam{\tau_1}{x}{\iota}}
+}
+
+\inferrule[syn-ap]{
+  \esynX{e_1}{\sty{\rightharpoonup}{(\st_1, \st_2)}}{\iota_1}\\\\
+  \eanaX{e_2}{\st_2}{\iota_2}
+}{
+  \esynX{\eap{e_1}{e_2}}{\st_2}{\iap{\iota_1}{\iota_2}}
+}
+
+\inferrule[ana-fix]{
+  \eana{\Upsilon, x \Rightarrow \st}{\Phi}{e}{\st}{\iota}\\\\
+  \sfinalrepX{\st}{\tau}
+}{
+  \eanaX{\efix{x}{e}}{\st}{\ifix{\tau}{x}{\iota}}
+}
+\end{mathpar}
+
+\subsubsection{Subsumption} Because we are defining a bidirectional type system, a subsumption rule is needed to allow synthetic terms to be analyzed against an equivalent type.
+Per above, equivalent types must be  syntactically identical, so the rule is straightforward:
+\begin{mathpar}\small
+\inferrule[subsume]{
+    \esynX{e}{\st}{\iota}
+}{
+    \eanaX{e}{\st}{\iota}
+}
+\end{mathpar}
+
+\subsubsection{Type Ascriptions}
+To use an analytic term in a synthetic position, the programmer must provide a type ascription, written $e : \st$.
+The ascription is kind checked and evaluated to a type before being used for analysis:
+\begin{mathpar}\small
+\inferrule[ascribe]{
+  \sofkz{\emptyset}{\emptyset}{\Phi}{\st}{\kty}\\
+  \snormT{\st}{\st'}\\
+  \eanaX{e}{\st'}{\iota}
+}{
+  \esynX{\easc{e}{\st}}{\st'}{\iota}
+}
+\end{mathpar}
+\subsubsection{Generalized Introductory Operations}
+The meaning of the generalized introductory form, $\eintro{\sttmidx}{\es}$, is determined by the tycon of the type it is being analyzed against as a function of the type's index, the \emph{term index}, $\sttmidx$, and the arguments, $\es$.
+
+Note that we can recover a variety of standard introductory forms (and allow their use at more than one type) by a purely syntactic desugaring to this form.
+The term index captures all static portions of the concrete form and the arguments capture all external sub-terms.
+Some example desugarings are defined below (using kind-specific syntax for static terms as above and writing $n$ and $\texttt{"s"}$ for static numbers and strings):
+\[\small
+\begin{array}{lll}
+\textbf{Description} ~~~~& \textbf{Concrete Form}~~~~ & \textbf{Desugared Form}\\
+\text{sequences} & (e_1, \ldots, e_n) ~\text{or}~ \texttt{[}e_1, \ldots, e_n\texttt{]} & \eintro{()}{e_1; \ldots; e_n}\\
+\text{labeled sequences} & \{\mathtt{lbl}_1=e_1, \ldots, \mathtt{lbl}_n=e_n\} & \eintro{\texttt{[}\mathtt{lbl}_1, \ldots, \mathtt{lbl}_n\texttt{]}}{e_1; \ldots; e_n}\\
+\text{label application} & \mathtt{lbl}\langle e_1, \ldots, e_n \rangle & \eintro{\mathtt{lbl}}{e_1, \ldots, e_n}\\
+\text{numerals} & n & \eintro{n}{\cdot}\\
+\text{labeled numerals} & n\texttt{lbl} & \eintro{(n, \texttt{lbl})}{\cdot}\\
+\text{strings} & \texttt{"s"} & \eintro{\texttt{"s"}}{\cdot}
+\end{array}
+\]
+
+As an example, we will derive $\small\eana{\Upsilon_\text{example}}{\Phi_\text{rstr}\Phi_\text{lprod}}{e_\text{example}}{\stx{paper}}{\iota_\text{example}}$ where the labeled product type $\stx{paper}$ and regular string type $\stx{title}$ was defined in Sec.
+\ref{types} and:\[\small\begin{array}{lcl}
+\Upsilon_\text{example} & := & title \Rightarrow \stx{title}\\
+e_\text{example} & := & \{\conclbl{title}=title, \conclbl{conf}=\texttt{"EXMPL 2015"}\}\\
+& \hookrightarrow & \eintro{\texttt{[}\conclbl{title}, \conclbl{conf}\texttt{]}}{title; \eintro{\texttt{"EXMPL 2015"}}{\cdot}}\\
+\iota_\text{example} & := & (title, (\texttt{"EXMPL 2015"}, ()))
+\end{array}
+\]
+
+The rule governing this form is below:
+\begin{mathpar}\small
+\inferrule[ana-intro]{
+  \tcdef{\tc}{\tcsig{\_}{\chi}}{\tcstruct{\_}{\omega}} \in \Phi\\\\
+  \introsig{\klitidx} \in \chi\\
+  \sofkz{\emptyset}{\emptyset}{\Phi}{\stmidx}{\klitidx}\\\\
+  \keyw{ana~intro}={\stx{def}} \in \omega\\
+  \keyw{args}(\es)=\stx{args}\\
+  \stx{def}~\sttyidx~\stmidx~\stx{args} \Downarrow_{\es;\Upsilon;\Phi} \sqitm{\qitm}\\\\
+  %\snorm{\stx{def}~\sttyidx~\stmidx~\stx{args}}{\emptyset}{\memG_0}{{\rightharpoonup}, \tc; \Upsilon; \Phi}{\sqitm{\itm_\text{abs}}}{\memD}{\memG}\\\\
+  %\snorm{\srep{\sty{\tc}{\sttyidx}}}{\memD}{\cdot}{{\rightharpoonup}, \tc;\emptyset;\Phi}{\sqity{\ity_\text{abs}}}{\memD'}{\cdot}\\\\
+  \edeabs{\tc}{{\es;\Upsilon;\Phi}}{\qitm}{\emptyset}{\emptyset}{\iota}{\memD}{\memG}\\
+  \tdeabs{\Phi}{\tc}{\srep{\sty{\tc}{\sttyidx}}}{\memD}{\ity}{\memD'}\\\\
+  \memD' \leadsto \delta : \Delta\\
+  \memG \leadsto \gamma : \Gamma\\
+  \Delta~\Gamma \vdash \iota : \tau
+}{
+  \eanaX{\eintro{\stmidx}{\es}}{\sty{\tc}{\sttyidx}}{[{\delta}][{\gamma}]\iota}
+}
+\end{mathpar}
+
+The first premise of (ana-intro) extracts the tycon definition for the tycon of the type the literal is being analyzed against, which is the {delegated tycon}, from $\Phi$.
+In this example, the delegated tycon is $\tcvar{lprod}$.
+
+The second premise extracts the \emph{intro index kind}, $\klitidx$, from the \emph{opcon signature}, $\chi$, of the delegated tycon's tycon signature.
+The third premise checks the provided term index against this kind.
+For example, $\tcvar{lprod}$ specifies $\klist{\klbl}$, so that it can use the labeled sequence form, while $\tcvar{rstr}$ specifies an intro index kind of $\keyw{Str}$, so that it can use the string literal form above:\[\small\begin{array}{lcl}
+\chi_\text{lprod} & := & \introsig{\klist{\klbl}}, \chi_\text{lprod/targops}\\
+\chi_\text{rstr} & := & \introsig{\keyw{Str}}, \chi_\text{rstr/targops}
+\end{array}
+\]
+
+The fourth premise extracts the \emph{intro opcon definition} from the \emph{opcon structure}, $\omega$, of the delegated tycon's tycon structure, calling it $\stx{def}$.
+This is the static function that is invoked to determine the translation of the term, returning a \emph{translational internal term} of kind $\kitm$ given the type index $\sttyidx$ of kind $\ktyidx$, the term index $\sttmidx$, of kind $\klitidx$, and an \emph{argument interface list}, $\stx{args}$, of kind $\kargs$, derived from $\es$ by the judgement $\keyw{args}(\es)=_n \st$, appearing as the fifth premise and described below.
+
+For example, the intro opcon definition for $\tcvar{rstr}$ is shown below:
+\[\begin{array}{lcl}
+\omega_\text{rstr} & := & \tcstructn{(
+    \slam{\krx}{\svar{tyidx}}{\slam{\kstr}{\svar{tmidx}}{\slam{\kargs}{\svar{args}}{\\
+&&\quad \keyw{let}~\svar{aok} :: \kunit = \svar{arity0}~\svar{args}~\keyw{in}~\\
+&&\quad \keyw{let}~\svar{rok} :: \kunit = \svar{rmatch}~\svar{tyidx}~\svar{tmidx}~\keyw{in}~\\
+&&\quad \sqitm{(\quq{\svar{str\_of\_Str}~\svar{tmidx}}, \quq{\svar{nat\_of\_Nat}~(\svar{len}~\svar{tmidx})})}
+}}}
+}),\\&&\omega_\text{rstr/targops}
+\end{array}\]
+
+The  judgement, $\vdash_\Phi \omega \sim \psi$, which appeared as the final premise of the rule (tcc-ext) above,  checks that such an intro opcon definition is well-kinded:
+\begin{mathpar}\small
+\inferrule[ocstruct-intro]{
+    \introsig{\klitidx} \in \chi\\
+    \emptyset \vdash \klitidx\\
+    \sofkz{\emptyset}{\emptyset}{\Phi}{\stx{def}}{\ktyidx \rightarrow \klitidx \rightarrow \kargs \rightarrow \kitm}
+}{
+    \vdash_\Phi \tcstructn{\stx{def}} \sim \tcsig{\ktyidx}{\chi}
+}
+\end{mathpar}
+
+This example intro opcon definition begins by making sure that no arguments were passed in, using the helper function $\svar{arity0} :: \kargs \rightarrow \kunit$ defined such that any non-empty input will raise an error, via the static term $\sraise{\kunit}$.
+The kinding rules and operational semantics of this exception-like form are standard:
+\begin{mathpar}\small
+\inferrule[k-raise]{\kDelta \vdash \kappa}{\sofkX{\sraise{\kappa}}{\kappa}}
+
+\inferrule[s-raise]{ }{\serr{\sraise{\kappa}}{\argEnv}}
+\end{mathpar}
+
+In practice, the tycon provider would specify an error message, but here we are satisfied simply signaling an error in this way.
+
+Next, this intro opcon definition checks the string provided as the term index against the regular expression given as the type index using $\svar{rmatch} :: \krx \rightarrow \keyw{Str} \rightarrow \kunit$, which we assume is defined in the usual way and again raises an error on failure.
+
+Finally, a \emph{translational internal term} corresponding to the term index is generated using the helper functions $\svar{str\_of\_Str} :: \keyw{Str} \rightarrow \kitm$ and $\svar{nat\_of\_Nat} :: \keyw{Nat} \rightarrow \kitm$ to generate  translational internal terms corresponding to the provided static terms and $\svar{len} :: \keyw{Str} \rightarrow \keyw{Nat}$ to statically compute the length of the string.
+
+The only introductory form for kind $\kitm$ is $\sqitm{\qitm}$, where $\qitm$ is a translational internal term.
+This form is analagous to the introductory form for kind $\kity$ described previously, $\sqity{\qity}$, where $\qity$ is a translational internal type.
+Each form in the syntax of $\iota$ has a corresponding form in the syntax for $\qitm$ and the kinding rules simply recurse through these generically, i.e.
+as shown for variables and lambda functions below:
+\begin{mathpar}\small
+\inferrule[k-itm-var]{ }{
+    \sofkX{\sqitm{x}}{\kitm}
+}
+
+\inferrule[k-itm-lam]{
+    \sofkX{\sqity{\qity}}{\kity}\\
+    \sofkX{\sqitm{\qitm}}{\kitm}
+}{
+    \sofkX{\sqitm{\ilam{\qity}{x}{\qitm}}}{\kitm}
+}
+\end{mathpar} 
+The operational semantics for these shared forms are similarly recursive:
+\begin{mathpar}\small
+\inferrule[s-itm-var-v]{ }{
+    \sval{\sqitm{x}}{\argEnv}
+}
+
+\inferrule[s-itm-lam-step-1]{
+    \sstep{\sqity{\qity}}{\argEnv}{\sqity{\qity'}}
+}{
+    \sstep{\sqitm{\ilam{\qity}{x}{\qitm}}}{\argEnv}{\sqitm{\ilam{\qity'}{x}{\qitm}}}
+}
+
+\inferrule[s-itm-lam-step-2]{
+    \sval{\sqity{\qity}}{\argEnv}\\
+    \sstep{\sqitm{\qitm}}{\argEnv}{\sqitm{\qitm'}}
+}{
+    \sstep{\sqitm{\ilam{\qity}{x}{\qitm}}}{\argEnv}{\sqitm{\ilam{\qity}{x}{\qitm'}}}
+}
+\end{mathpar}
+\begin{mathpar}\small
+\inferrule[s-itm-lam-err-1]{
+    \serr{\sqity{\qity}}{\argEnv}
+}{
+    \serr{\sqitm{\ilam{\qity}{x}{\qitm}}}{\argEnv}
+}
+
+\inferrule[s-itm-lam-err-2]{
+    \serr{\sqitm{\qitm}}{\argEnv}
+}{
+    \serr{\sqitm{\ilam{\qity}{x}{\qitm}}}{\argEnv}
+}
+
+\inferrule[s-itm-lam-v]{
+    \sval{\sqity{\qity}}{\argEnv}\\
+    \sval{\sqitm{\qitm}}{\argEnv}
+}{
+    \sval{\sqitm{\ilam{\qity}{x}{\qitm}}}{\argEnv}
+}
+\end{mathpar}
+Like translational internal types, there is also an unquote form, $\quq{\st}$, that permits translational internal terms to be constructed compositionally:
+\begin{mathpar}\small
+\inferrule[k-itm-unquote]{
+    \sofkX{\st}{\kitm}
+}{
+    \sofkX{\sqitm{\quq{\st}}}{\kitm}
+}
+\end{mathpar}
+These unquote forms are eliminated in the course of evaluation:
+\begin{mathpar}\small
+\inferrule[s-itm-unquote-step]{
+    \sstep{\st}{\argEnv}{\st'}
+}{
+    \sstep{\sqitm{\quq{\st}}}{\argEnv}{\sqitm{\quq{\st'}}}
+}
+
+\inferrule[s-itm-unquote-err]{
+    \serr{\st}{\argEnv}
+}{
+    \serr{\sqitm{\quq{\st}}}{\argEnv}
+}
+
+\inferrule[s-itm-unquote-elim]{
+    \sval{\sqitm{\qitm}}{\argEnv}
+}{
+    \sstep{\sqitm{\quq{\sqitm{\qitm}}}}{\argEnv}{\sqitm{\qitm}}
+}
+\end{mathpar}
+
+The final two forms of translational internal term are $\anatrans{n}{\st}$ and $\syntrans{n}$.
+These represent the translation of argument $n$, the first if it arises via analysis against type $\st$ and the second if it arises by synthesis.
+They are not intended to be written directly by tycon providers, arising only internally in the semantics of argument interface lists.
+Before giving the rules, let us motivate the mechanism by showing the intro opcon definition for $\tcvar{lprod}$:
+\[
+\begin{array}{lcl}
+\omega_\text{lprod} & := & \tcstructn{\lambda\svar{tyidx}{:}\klist{\klbl\times\kty}.\lambda\svar{tmidx}{:}\klist{\klbl}.\lambda\svar{args}{:}\kargs.\\
+&&\quad \keyw{let}~\svar{inhabited}:\kunit=\svar{uniqmap}~\svar{tyidx}\\
+&&\quad \skap{\kitm}{\skap{\karg}{\skap{\klbl}{\skap{\klbl\times\kty}{\svar{listrec3}}}}}}~\svar{tyidx}~\svar{tmidx}~\svar{args}~\sqitm{\itriv}\\
+&&\quad\quad \lambda \svar{rowtyidx}{:}\klbl\times\kty.\lambda \svar{rowtmidx}{:}\klbl.\lambda\svar{rowarg}{:}\karg.\lambda \svar{r}{:}\kitm.\\
+&&\quad\quad\quad \keyw{letpair}~(\svar{rowlbl}, \svar{rowty})=\svar{rowtyidx}\\
+&&\quad\quad\quad \keyw{let}~\svar{lok}::\kunit=\svar{lbleq}~\svar{rowlbl} ~\svar{rowtmidx}\\
+&&\quad\quad\quad \keyw{let}~\svar{rowtr} :: \kitm = \svar{ana}~\svar{rowarg}~\svar{rowty}\\
+&&\quad\quad\quad \sqitm{(\quq{\svar{rowtr}}, \quq{\svar{r}})
+},\\
+&&\omega_\text{lprod/targops}
+\end{array}
+\]
+
+The first line of the intro opcon definition for $\tcvar{lprod}$ checks that the type index is well-formed, in this case by checking that there are no duplicate labels via the helper function $\svar{uniqmap} :: \klist{\klbl \times \kty}\rightarrow \kunit$, raising an error if there are.
+An alternative strategy may have been to use an abstract kind that ensured that such type indices could not have been constructed, but to be compatible with our equality kind restriction, this would require support for abstract equality kinds, analagous to abstract equality types in SML.
+We chose not to formalize these for simplicity, and to demonstrate this more general technique: types with indices that are to be considered ``malformed'' can simply be left uninhabited.
+An analagous technique could be used to implement record types, leaving indices where the labels did not appear sorted uninhabited.
+In both cases, we could provide a static helper function of kind $\klist{\klbl\times\kty}\rightarrow \kty$ that checked well-formedness immediately before constructing the requested type.
+
+The rest of this intro opcon definition folds over the three lists provided as input: the list mapping labels to types provided as the type index, the list of labels provided as the term index, and the list of argument interfaces.
+We assume a straightforward helper function, $\svar{listfold3}$, that raises an error if the three lists are not of the same length.
+
+The base case is the translational empty product.
+The recursive case first checks that the label provided in the term index matches the label provided in the type index, using a helper function $\svar{lbleq} :: \klbl \rightarrow \klbl \rightarrow \kunit$.
+Then, we request type analysis of the corresponding argument, $\svar{rowarg}$,  against the type in the type index, $\svar{rowty}$, by writing $\svar{ana}~\svar{rowarg}~\svar{rowty}$.
+
+We define $\karg$, the kind of \emph{argument interfaces}, as a product of functions, one for analysis and the other for synthesis, and the helper functions $\svar{ana}$ and $\svar{syn}$ simply project the corresponding function out:\todo{fix products}
+\[\begin{array}{lcl}
+\karg & := & (\kty \rightarrow \kitm) \times (\kunit \rightarrow \kty\times\kitm)\\
+\svar{ana} & := & \slam{\karg}{\svar{arg}}{\keyw{fst}(\svar{arg})}\\
+\svar{syn} & := & \slam{\karg}{\svar{arg}}{\keyw{snd}(\svar{arg})}
+\end{array}
+\]
+
+%The $\kargs$ given where the $n$th entry is simply a pair of functions, the first of which allows for analysis of the $n$th argument of the operation against a type, and the second of which requests type synthesis for the $n$th argument.
+The helper functions simply project out the appropriate functions and invoke them.
+
+As mentioned above, the argument interface list is constructed from the argument list by the judgement $\keyw{args}(\es)=_n \stx{args}$, defined as follows:
+\begin{mathpar}\small
+\inferrule[args-z]{ }{
+    \keyw{args}(\cdot)=_0 \skap{\karg}{\svar{nil}}
+}
+
+\inferrule[args-s]{
+    \keyw{args}(\es)=_{n} \st
+}{
+    \keyw{args}(\es;e)=_{n+1} \skap{\karg}{\svar{rcons}}~\st~(\slam{\kty}{\svar{ty}}{\sana{n}{\svar{ty}}},\slam{\kunit}{\_}{\ssyn{n}})
+}
+\end{mathpar}
+We assume that the definitions of the standard helper functions $\svar{nil} :: \kforall{\kalpha}{\klist{\kalpha}}$ and $\svar{rcons} :: \kforall{\kalpha}{\klist{\kalpha} \rightarrow \kalpha \rightarrow \klist{\kalpha}}$, which adds an item to the end of a list, have been substituted into these rules.
+
+The $n$th element of the argument interface list simply wraps the static terms $\sana{n}{\st}$ and $\ssyn{n}$.
+Recall that the kinding judgement is indexed by $n$, which is an upper bound on the argument index of such terms.
+This is enforced in the kinding rules:
+\begin{mathpar}\small
+\inferrule[k-ana]{
+    n' < n\\
+    \sofkX{\st}{\kty}
+}{
+    \sofkX{\sana{n'}{\st}}{\kitm}
+}
+
+\inferrule[k-syn]{
+    n' < n
+}{
+    \sofkX{\ssyn{n'}}{\kty\times\kitm}
+}
+\end{mathpar}
+
+The following lemma thus characterizes the argument interface list judgement:
+\begin{lemma}
+If $\keyw{args}(\es)=_n \st$ then $\sofkn{\emptyset}{\emptyset}{\Phi}{n}{\st}{\kargs}$.
+\end{lemma}
+
+The rule (ocstruct-intro) ruled out writing either of these forms explicitly in an opcon definition by checking against the bound $n=0$.
+But accessing them via the argument interface list is possible because the following straightforward lemma holds (see appendix for proofs):
+\begin{lemma}
+If $\sofkn{\kDelta}{\kGamma}{\Phi}{n'}{\st}{\kappa}$ and $n > n'
+$ then $\sofkn{\kDelta}{\kGamma}{\Phi}{n}{\st}{\kappa}$.
+\end{lemma}
+
+The operational semantics for these couple the dynamics of the static language to the statics of the external language.
+For $\sana{n}{\st}$, after stepping the type $\st$ to a value and propagating errors, the argument environment, which stores the arguments themselves and the typing and tycon contexts, $\es; \Upsilon; \Phi$, is consulted to retrieve the $n$th argument and analyze it against $\st$.
+If this succeeds, the translational internal term $\sqitm{\anatrans{n}{\st}}$ is generated to refer to it.
+If it fails, an error is raised.
+We write the judgement $[\Upsilon \vdash_\Phi e \nLeftarrow \st]$ to indicate that $e$ fails to analyze against $\st$.
+Because we do not here define this judgement itself inductively, we also allow that this premise be left out entirely, resulting in a non-deterministic but sufficient semantics for our metatheory:
+\begin{mathpar}\small
+\inferrule[s-ana-step]{
+    \sstep{\st}{\argEnv}{\st'}
+}{
+    \sstep{\sana{n}{\st}}{\argEnv}{\sana{n}{\st'}}
+}
+
+\inferrule[s-ana-success]{
+    \svalNA{\st}\\
+    \keyw{nth}[n](\es) = e\\
+    \eana{\Upsilon}{\Phi}{e}{\st}{\iota}
+}{
+    \sstep{\sana{n}{\st}}{\es;\Upsilon;\Phi}{\sqitm{\anatrans{n}{\st}}}
+}
+
+\inferrule[s-ana-err]{
+    \serr{\st}{\argEnv}
+}{
+    \serr{\sana{n}{\st}}{\argEnv}
+}
+
+\inferrule[s-ana-fail]{
+    \svalNA{\st}\\
+    \keyw{nth}[n](\es) = e\\
+    [\Upsilon \vdash_\Phi e \nLeftarrow \st]
+}{
+    \serr{\sana{n}{\st}}{\es;\Upsilon;\Phi}
+}
+\end{mathpar}
+The semantics for $\ssyn{n}$ are analagous, evaluating to a pair containing the translational internal term $\sqitm{\syntrans{n}}$ as well as the synthesized type:
+\begin{mathpar}\small
+\inferrule[s-syn-success]{
+    \keyw{nth}[n](\es) = e\\
+    \esynX{e}{\st}{\iota}
+}{
+    \sstep{\ssyn{n}}{\es;\Upsilon;\Phi}{(\st, \sqitm{\syntrans{n}})}
+}
+
+\inferrule[s-syn-fail]{
+    \keyw{nth}[n](\es) = e\\
+    [\Upsilon \vdash_\Phi e \nRightarrow]
+}{
+    \serr{\ssyn{n}}{\es;\Upsilon;\Phi}
+}
+\end{mathpar}
+
+The kinding rules also prevent the translational internal term forms generated by these operations from being written directly when $n = 0$:
+\begin{mathpar}\small
+\inferrule[k-itm-anatrans]{
+    n' < n\\
+    \sofkX{\st}{\kty}
+}{
+    \sofkX{\sqitm{\anatrans{n'}{\st}}}{\kitm}
+}
+
+\inferrule[k-itm-syntrans]{
+    n' < n
+}{
+    \sofkX{\sqitm{\syntrans{n'}}}{\kitm}
+}
+\end{mathpar}
+Like the translational internal type form $\srep{\st}$, these two translational internal term forms are retained in values of kind $\kitm$:
+\begin{mathpar}\small
+\inferrule[s-itm-anatrans-step]{
+    \sstep{\st}{\argEnv}{\st'}
+}{
+    \sstep{\sqitm{\anatrans{n}{\st}}}{\argEnv}{\sqitm{\anatrans{n}{\st'}}}
+}
+
+\inferrule[s-itm-anatrans-err]{
+    \serr{\st}{\argEnv}
+}{
+    \serr{\sqitm{\anatrans{n}{\st}}}{\argEnv}
+}
+
+\inferrule[s-itm-anatrans-v]{
+    \svalNA{\st}\\
+    \keyw{nth}[n](\es)=e
+}{
+    \sval{\sqitm{\anatrans{n'}{\st}}}{\es;\Upsilon;\Phi}
+}
+
+\inferrule[s-itm-syntrans-v]{
+    \keyw{nth}[n](\es)=e
+}{
+    \sval{\sqitm{\syntrans{n}}}{\es;\Upsilon;\Phi}
+}
+\end{mathpar}
+
+The final line of the intro opcon definition for $\tcvar{lprod}$ constructs a nested tuple as the translation.
+Taken together, the translational internal term that will be derived for our example involving $e_\text{example}$ above is:
+\[\begin{array}{lcl}
+\qitm_\text{example} & := & (\anatrans{0}{\stx{title}}, (\anatrans{1}{\stx{conf}}, ()))
+\end{array}
+\]
+
+The reason that the translations themselves were not inserted yet (even though they were derived in rule (s-ana-success) above) is again because we want to be able to selectively hold these abstract.
+The selective term abstraction judgement $\edeabs{\tc}{\argEnv}{\qitm}{\memD}{\memG}{\itm}{\memD'}{\memG'}$, appearing as the seventh premise of (ana-intro), relates a translational internal term $\qitm$ to an internal term $\itm$ called the corresponding abstract internal term, where all references to the translation of a type constructed by a user-defined tycon other than the ``delegated tycon'' $\tc$ are replaced with an abstract type variable, and all references to the translation of an argument (having any type) are replaced with a variable.
+The type translation store, $\memD$, discussed previously, and term translation store, $\memG$, track these mappings.
+Term translation stores have the following syntax:\[
+\begin{array}{lcl}
+\memG & ::= & \emptyset ~|~ \memG, n : \st \leadsto \iota/x : \tau
+\end{array}\]
+
+Each entry can be read ``argument $n$ having type $\st$ and translation $\iota$ appears as variable $x$ with type $\tau$''.
+For example, we have that \[\edeabs{\tcvar{lprod}}{\es,\Upsilon,\Phi_\text{rstr}\Phi_\text{lprod}}{\qitm_\text{example}}{\emptyset}{\emptyset}{\iota_\text{example/abs}}{\memD_\text{example/abs}}{\memG_\text{example/abs}}\] where \[
+\begin{array}{lcl}
+\itm_\text{example/abs} & := & (x_0, (x_1, ()))\\
+\memG_\text{example/abs} & := & 0 : \stx{title} \leadsto title/x_0 : \alpha_0, 1 : \stx{conf} \leadsto \texttt{"EXMPL 2015"}/x_1 : \alpha_1\\
+\memD_\text{example/abs} & := & \stx{title} \leftrightarrow \keyw{str}/\alpha_1, \stx{conf} \leftrightarrow \keyw{str}/\alpha_2
+\end{array}\]
+The rules for the selective term abstraction judgement follow recursively for shared forms, e.g. for variables and lambdas:
+\begin{mathpar}\small
+\inferrule[abs-var]{ }{\edeabs{\tc}{\argEnv}{x}{\memD}{\memG}{x}{\memD}{\memG}}
+
+\inferrule[abs-lam]{
+    \tdeabs{\Phi}{\tc}{\qity}{\memD}{\ity}{\memD'}\\
+    \edeabs{\tc}{\es;\Upsilon;\Phi}{\qitm}{\memD'}{\memG}{\itm}{\memD''}{\memG'}
+}{
+    \edeabs{\tc}{\es;\Upsilon;\Phi}{\ilam{\qity}{x}{\qitm}}{\memD}{\memG}{\ilam{\tau}{x}{\itm}}{\memG'}{\memD''}
+}
+\end{mathpar}
+The rules for references to argument translations operate as follows:
+\begin{mathpar}\small
+\inferrule[abs-anatrans-stored]{
+    n : \st \leadsto \iota/x : \tau \in \memG
+}{
+    \edeabs{\tc}{\argEnv}{\anatrans{n}{\st}}{\memG}{\memD}{x}{\memG}{\memD}
+}
+
+\inferrule[abs-syntrans-stored]{
+    n : \st \leadsto \iota/x : \tau \in \memG
+}{
+    \edeabs{\tc}{\argEnv}{\syntrans{n}}{\memG}{\memD}{x}{\memG}{\memD}
+}\end{mathpar}
+\begin{mathpar}\small
+\inferrule[abs-anatrans-new]{
+    n \notin \text{dom}(\memG)\\
+    \keyw{nth}[n](\es) = e\\
+    \eanaX{e}{\st}{\iota}\\
+    \tdeabs{\Phi}{\tc}{\srep{\st}}{\memD}{\ity}{\memD'}\\
+    (x~\text{fresh})
+}{
+    \edeabs{\tc}{\es;\Upsilon;\Phi}{\anatrans{n}{\st}}{\memG}{\memD}{x}{\memG, n : \st \leadsto \iota/x : \tau}{\memD'}
+}
+\end{mathpar}
+\begin{mathpar}\small
+\inferrule[abs-syntrans-new]{
+    n \notin \text{dom}(\memG)\\
+    \keyw{nth}[n](\es) = e\\
+    \esynX{e}{\st}{\iota}\\
+    \tdeabs{\Phi}{\tc}{\srep{\st}}{\memD}{\ity}{\memD'}\\
+    (x~\text{fresh})
+}{
+    \edeabs{\tc}{\es;\Upsilon;\Phi}{\syntrans{n}}{\memG}{\memD}{x}{\memG, n : \st \leadsto \iota/x : \tau}{\memD'}
+}
+\end{mathpar}
+
+After computing an abstract term translation, the eight premise computes an abstract type translation for the type the intro term is being analyzed against:
+\[\small\tdeabs{\Phi_\text{rstr}\Phi_\text{lprod}}{\tcvar{lprod}}{\srep{\stx{conf}}}{\memD_\text{example/abs}}{\tau_\text{example/abs}}{\memD_\text{example/abs}}\]
+where $\tau_\text{example/abs} := \alpha_0 \times (\alpha_1 \times \kunit)$ (alpha-equivalent to $\tau_\text{conf/abstrans}$ in Sec. \ref{types}).
+
+Each term translation store $\memG$ induces an internal term substitution, written in the standard way $\gamma ::= \emptyset ~|~ \gamma, \iota/x$, and a corresponding internal typing context $\Gamma$, as specified by the judgement $\memG \leadsto \gamma : \Gamma$ defined by the following rules:
+\begin{mathpar}
+\inferrule[ttrs-emp]{ }{\emptyset \leadsto \emptyset : \emptyset}
+
+\inferrule[ttrs-ext]{
+    \memG \leadsto \gamma : \Gamma
+}{
+    (\memG, n : \st \leadsto \iota/x : \tau) \leadsto (\gamma, \iota/x) : (\Gamma, x : \tau)
+}
+\end{mathpar}
+In this case, we have the ninth premise, $\memG_\text{example/abs} \leadsto \gamma_\text{example/abs} : \Gamma_\text{example/abs}$, where \[
+\begin{array}{lcl}
+\gamma_\text{example/abs} & := & title/x_0, \texttt{"EXMPL 2015"}/x_1\\
+\Gamma_\text{example/abs} & := & x_0 : \alpha_0, x_1 : \alpha_1
+\end{array}
+\]
+
+The tenth premise extracts a type substitution, $\delta_\text{example/abs}$, and type formation context, $\Delta_\text{example/abs}$, from $\memD_\text{example/abs}$, again equivalent to those defined in Sec.
+\ref{types}.
+
+Finally, the tenth premise checks that the abstract term translation is consistent with the abstract type translation: $\Delta_\text{example/abs}~\Gamma_\text{example/abs} \vdash \iota_\text{example/abs} : \tau_\text{example/abs}$, i.e.
+\[
+(\alpha_0, \alpha_1)~(x_0 : \alpha_0, x_1 : \alpha_1) \vdash (x_0, (x_1, ())) : \alpha_0 \times (\alpha_1 \times \kunit)
+\]
+
+Observe that the translation of the labeled product $e_\text{example}$ containing regular strings is being checked with all references to term and type translations of regular strings replaced by variables and type variables, respectively.
+Nevertheless, because our logic for labeled products has been treating regular strings parametrically, the check succeeds.
+
+Applying the substitutions $\gamma_\text{example/abs}$ and $\delta_\text{example/abs}$ in the conclusion of the rule, we have our final translation, $\iota_\text{example}$, i.e.
+$(title, (\texttt{"EXMPL 2015"}, ())$.
+
+Note that $\Gamma_\text{example} \vdash \iota_\text{example} : \tau_\text{conf}$, where $\vdash_{\Phi_\text{rstr}\Phi_\text{lprod}} \Upsilon_\text{example} \leadsto \Gamma_\text{example}$ and $\vdash_{\Phi_\text{rstr}\Phi_\text{lprod}} \sigma_\text{conf} \leadsto \tau_\text{conf}$ as described in Sec.
+\ref{types}.
+In fact, this relationship between the translation of an external term and the translation of the type it has will always hold given the check we performed on the abstract term and type transations just described due to parametricity properties.
+Combined with type safety of the IL this will imply type safety of @$\lambda$.
+We give the details in Sec.
+\ref{metatheory}.
+
+Had the translational term generated by the intro opcon definition been, e.g., $(\anatrans{0}{\stx{title}}, (\texttt{"TEST"}, ())$, then the corresponding abstract term would be $(x_0, (\texttt{"TEST"}, ())$ and the check would fail because $(\alpha_0, \alpha_1)~(x_0 : \alpha_0) \nvdash \texttt{"TEST"} : \alpha_1$.
+Had we not gone through the machinations of holding types and terms abstract, however, the check would succeed, because $\keyw{str}$ would no longer have been held abstract as $\alpha_1$.
+This form of ``type translation independence'' is precisely analagous to the representation independence properties that underly abstraction theorems for module systems based on  abstract types and will similarly serve to ensure that the type invariants maintained by $\tcvar{rstr}$ are conserved when new tycons are defined.
+In this case the invariant is that only strings that are in the regular language specified in the type index can be generated from an external term of regular string type.
+Note that $\texttt{"TEST"}$ is not in the regular language specified by $\concrx{[A-Z]+ \digit\digit\digit\digit}$, so it is critical that it not be allowed, despite being consistent with the type translation of $\stx{conf}$, i.e.
+it is a string.
+Such an invariant could not be maintained if the type system were treated as merely a ``bag of rules''.
+We consider this more rigorously in Sec.
+\ref{metatheory}.
+
+Note also that the domains of $\Upsilon_\text{example}$ and $\Gamma_\text{example/abs}$ are disjoint.
+This serves to ensure \emph{hygienic translation} -- translations cannot refer to variables in the surrounding scope directly, so uniformly renaming a variable cannot change the meaning of a program.
+Variables in $\Upsilon_\text{example}$ can (and do -- $title$ in this example) occur in arguments to the operation, but the translations of the arguments only appear after the substitution $\gamma_\text{example/abs}$ has been applied.
+We assume that applying a substitution is capture-avoiding in the usual manner (i.e.
+consistent with a locally nameless implementation).
+\subsubsection{Generalized Targeted Operations} 
+All non-introductory opcons associated with user-defined tycons are used via a form for \emph{targeted operations}, $\etarg{\opname{op}}{\sttmidx}{e_\text{targ}}{\es}$, where $\opname{op}$ is the opcon name, ${\sttmidx}$ is the term index, $e_\text{targ}$ is the \emph{target argument} and $\es$ are the remaining arguments.
+Some examples of desugarings to this form are below:
+\[\small
+\begin{array}{lll}
+\textbf{Description}& \textbf{Concrete Form}~~~~ & \textbf{Desugared Form}\\
+\text{index projection} & e_\text{targ}\texttt{\#}n & \etarg{\opname{idx}}{n}{e_\text{targ}}{\cdot}\\
+\text{label projection} & e_\text{targ}\texttt{\#}\conclbl{lbl} & \etarg{\opname{prj}}{\conclbl{lbl}}{e_\text{targ}}{\cdot}\\
+\text{explicit invocation}~~~~ & e_\text{targ}{\cdot}\opname{op}[\sttmidx](\es) & \etarg{\opname{op}}{\sttmidx}{e_\text{targ}}{\es}\\
+& e_\text{targ}{\cdot}\opname{op}(\es) & \etarg{\opname{op}}{()}{e_\text{targ}}{\es}\\
+& e_\text{targ}{\cdot}\opname{op}(\conclbl{lbl}_1=e_1, \ldots, \conclbl{lbl}_n=e_n) & \keyw{targ}[\opname{op}; \texttt{[}\conclbl{lbl}_1, \ldots, \conclbl{lbl}_n\texttt{]}](e_\text{targ}; \\
+& & \quad e_1; \ldots; e_n)\\
+\text{labeled case analysis} & e_\text{targ}{\cdot}\opname{case}\,\{ & \keyw{targ}[{\opname{case}}; {\texttt{[}\st_1, \ldots, \st_n\texttt{]}}]({e_\text{targ}}; \\
+ & ~|~~\st_1\langle x_1, \ldots, x_k \rangle \Rightarrow e_1 & \quad \eanalam{x_1}{\ldots \eanalam{x_k}{e_1}};\\
+ & ~|~~\ldots & \quad \ldots;\\ 
+ & ~|~~\st_n\langle x_1, \ldots, x_k \rangle \Rightarrow e_n\} & \quad \eanalam{x_1}{\ldots \eanalam{x_k}{e_n}})
+\end{array}
+\]
+
+Whereas introductory operations were strictly analytic, targeted operations are always synthetic in @$\lambda$.
+The type and translation is determined by the tycon of the type synthesized by the target argument (i.e.
+this tycon is used as the delegated tycon).
+The rule given below is otherwise similar to (ana-intro) in its key machinations:
+\begin{mathpar}\small
+\inferrule[syn-targ]{
+  \esynX{e_\text{targ}}{\sty{\tc}{\sttyidx}}{\iota_\text{targ}}\\
+  \tcdef{\tc}{\tcsig{\_}{\chi}}{\tcstruct{\_}{\omega}} \in \Phi\\\\
+  \opsig{\opname{op}}{\klitidx} \in \chi\\
+  \sofkz{\emptyset}{\emptyset}{\Phi}{\stmidx}{\klitidx}\\\\
+  \keyw{syn~}\opname{op}={\stx{def}} \in \omega\\
+  \keyw{args}(e_\text{targ}; \es)=\stx{args}\\
+  \stx{def}~\sttyidx~\stmidx~\stx{args} \Downarrow_{(e_\text{targ}; \es);\Upsilon;\Phi} (\st, \sqitm{\qitm})\\\\
+  %\snorm{\stx{def}~\sttyidx~\stmidx~\stx{args}}{\emptyset}{\memG_0}{{\rightharpoonup}, \tc; \Upsilon; \Phi}{\sqitm{\itm_\text{abs}}}{\memD}{\memG}\\\\
+  %\snorm{\srep{\sty{\tc}{\sttyidx}}}{\memD}{\cdot}{{\rightharpoonup}, \tc;\emptyset;\Phi}{\sqity{\ity_\text{abs}}}{\memD'}{\cdot}\\\\
+  \edeabs{\tc}{{\es;\Upsilon;\Phi}}{\qitm}{\emptyset}{\emptyset}{\iota}{\memD}{\memG}\\
+  \tdeabs{\Phi}{\tc}{\srep{\st}}{\memD}{\ity}{\memD'}\\\\
+  \memD' \leadsto \delta : \Delta\\
+  \memG \leadsto \gamma : \Gamma\\
+  \Delta~\Gamma \vdash \iota : \tau
+}{
+    \esynX{\etarg{\opname{op}}{\sttmidx}{e_\text{targ}}{\es}}{\st}{[\delta][\gamma]\iota}
+}
+\end{mathpar}
+
+The first premise synthesizes a type, $\sty{\tc}{\sttyidx}$, for the target argument.
+The second premise extracts the tycon definition for $\tc$ from the tycon context.
+The third extracts the \emph{operator index kind} from its opcon signature, and the fourth checks the term index against this kind.
+For example, we may associate the following targeted opcon signatures with the tycons $\tcvar{rstr}$ and $\tcvar{lprod}$:\[
+\begin{array}{lcl}
+\chi_\text{rstr/targops} & := & \opsig{\opname{concat}}{\kunit}, \opsig{\opname{case}}{\klist{\keyw{StrPattern}}}, \opsig{\opname{coerce}}{\krx}, \opsig{\opname{check}}{\krx}, \opsig{\opname{replace}}{\krx}\\
+\chi_\text{lprod/targops} & := & \opsig{\opname{prj}}{\klbl}, \opsig{\opname{with}}{\klist{\klbl}}, \opsig{\opname{drop}}{\klist{\klbl}}
+\end{array}
+\]
+
+The opcons associated with $\tcvar{rstr}$ are taken directly from Fulton et al.'s specification of regular string types \cite{sanitation-psp14}, with the exception of $\opname{case}$, which generalizes the case analysis operator given there to arbitrary string patterns, which we assume are encoded in some suitable way by kind $\keyw{StrPattern}$.
+The opcons associated with $\tcvar{lprod}$ are also straightforward: $\opname{prj}$ projects out a labeled value from a labeled product, $\opname{with}$ creates a new labeled product based on the target with some fields updated or added, and $\opname{drop}$ creates a new labeled product from the target with some fields dropped.
+Note that regular string types cannot directly be embedded into ML-like languages and the analogs of $\opname{with}$ and $\opname{drop}$ are only found in some dialects of ML.
+For concision, we will only show $\opname{concat}$ and $\opname{prj}$ here; more details on the others can be found in the appendix.
+
+The fifth premise of (syn-targ) extracts the targeted opcon definition of $\opname{op}$ from the opcon structure, $\omega$.
+Like the intro opcon definition, this is a static function that generates a translational internal term on the basis of the delegated tycon's type index, the term index and an argument interface list.
+Targeted opcon definitions pair the translation with the synthesized type as well.
+The rule (ocstruct-op) ensures that the opcon definition has a correct and well-formed kind:
+\begin{mathpar}\small
+\inferrule[ocstruct-op]{
+    \vdash_\Phi \omega \sim \tcsig{\ktyidx}{\chi}\\
+    \opname{op}\notin\text{dom}(\chi)\\
+    \emptyset \vdash \klitidx\\\\
+    \sofkn{\emptyset}{\emptyset}{\Phi}{0}{\stx{def}}{\ktyidx \rightarrow \klitidx \rightarrow \kargs \rightarrow (\kty \times \kitm)}
+}{
+    \vdash_\Phi \tcstructc{\omega}{op}{\stx{def}} \sim \tcsig{\ktyidx}{\chi, \opsig{\opname{op}}{\klitidx}}
+}
+\end{mathpar}
+
+For example, the opcon definition for $\opname{concat}$ is defined as follows:
+\[\small
+\begin{array}{lcl}
+\omega_\text{rstr/targops} & := & \keyw{syn}~\opname{concat}=\slam{\krx}{\svar{tyidx}}{
+    \slam{\kunit}{\svar{tmidx}}{
+        \slam{\kargs}{\svar{args}}{\\&&
+            \quad \keyw{letpair}~(\svar{arg1}, \svar{arg2}) = \svar{arity2}~\svar{args}\\&&
+            \quad \keyw{letpair}~(\_, \svar{tr1}) = \svar{syn}~\svar{arg1}\\&&
+            \quad \keyw{letpair}~(\svar{ty2}, \svar{tr2}) = \svar{syn}~\svar{arg2}\\&&
+            \quad \stycase{\tcvar{rstr}}{\svar{ty2}}{\svar{tyidx2}}{\\&&
+                \quad\quad (\sty{\tcvar{rstr}}{\svar{rxconcat}~\svar{tyidx}~\svar{tyidx2}}, \sqitm{sconcat~\quq{\svar{tr1}}~\quq{\svar{tr2}}})\\&&\quad\quad
+            }{\sraise{\kty\times\kitm}}
+        }
+    }
+}
+\end{array}
+\]
+The first line simply checks that exactly two total arguments, including the target term, were passed in using the helper function $\svar{arity2}$.
+We then request synthesis of both arguments.
+We can ignore the type synthesized by the first because by definition it is a regular string type with type index $\svar{tyidx}$.
+We case analyze the second against $\tcvar{rstr}$, extracting its index regular expression if so and failing if not.
+We finally synthesize the resulting regular string type, using the helper function $\svar{rxconcat} :: \krx \rightarrow \krx \rightarrow \krx$ which generates the concatenated regular expression, and the translation, using an internal helper function $sconcat : \keyw{str} \rightarrow \keyw{str} \rightarrow \keyw{str}$, the translational internal term for which we assume has been substituted in directly above.
+
+This definition is invoked as the sixth premise of (syn-targ), generating the type ...
+and the translational internal term ...
+consistent with the rules described above.
+The remaining premises are identical to the corresponding premises in (ana-intro), with the only exception being that we check the abstract term translation against the abstract type translation of ...
+In this case...
+Note that because the delegated tycon and the tycon of the synthesized type are the same, the type is not held abstract, allowing our use of the helper function $sconcat$ to successfully typecheck.
+Had an identical translational internal term arisen from an opcon associated with $\tcvar{lprod}$, the corresponding abstract type translation would be a type variable, $\alpha$, so the check would fail.
+This implies that the choice of representation, and of representation invariants, for regular strings is entirely at the disrection of the $\tcvar{rstr}$ definition.
+More creative possibilities could be used with only local impact, e.g.
+traces through the regular language or ...
+
+\subsubsection{Other} The final form of external term is $\eother{m}{\iota}$.
+It can be analyzed against $\sotherty{m}{\tau}$ if $\iota$ has type $\tau$ in the translation of the current typing context.
+We will see how it is used as a technical device in the next section.
+\begin{mathpar}\small
+\inferrule[ana-other]{
+    \vdash_\Phi \Upsilon \leadsto \Gamma\\
+    \emptyset~\Gamma \vdash \iota : \tau
+}{
+    \eanaX{\eother{\iota}}{\sotherty{m}{\tau}}{\iota}
+}
+\end{mathpar}
+
+\section{Metatheory}\label{metatheory}
+%This judgement is only defined for values of kind $\kty$. We write $\st~\texttt{type}_\Phi$ iff $\vdash \Phi$ and $\sofkn{\emptyset}{\emptyset}{\Phi}{0}{\st}{\kty}$ and $\sval{\st}$.
+
+\begin{definition}[Argument Environment Formation] $\argEnvOK{n}{\es;\Upsilon;\Phi}$ iff $|\es|=n$ and $\vdash \Phi$ and $\vdash_\Phi \Upsilon \leadsto \Gamma$.\end{definition}
+
+\subsection{Static Language}
+
+%\begin{lemma}
+%If $\kDelta \vdash \kGamma$ and $\vdash \Phi$ and $\sofkn{\kDelta}{\kGamma}{\Phi}{n}{\st}{\kappa}$ then $\kDelta \vdash \kappa$.\todo{don't think we need this}
+%\end{lemma}
+
+\begin{theorem}[Static Canonical Forms] If $\sofkn{\emptyset}{\emptyset}{\Phi}{n}{\st}{\kappa}$ and $\argEnvOK{n}{\argEnv}$ and $\sval{\st}{\argEnv}$ then:
+\begin{enumerate}
+\item TODO: arrow
+\item TODO: unit
+\item TODO: product
+\item TODO: sum
+\item TODO: inductive
+\item TODO: universal
+\item If $\kappa = \kty$ then $\istype{\st}{\Phi}$ and 
+    \begin{enumerate}
+    \item $\st = \sty{\rightharpoonup}{(\st_1, \st_2)}$ and $\istype{\st_1}{\Phi}$ and $\istype{\st_2}{\Phi}$; or
+    \item $\st = \sty{\tc}{\sttyidx}$ and $\tcdef{\tc}{\tcsig{\ktyidx}{\chi}}{\omega} \in \Phi$ and $\sofkn{\emptyset}{\emptyset}{\Phi}{0}{\st}{\ktyidx}$ and $\svalNA{\sttyidx}$; or
+    \item $\st = \sotherty{m}{\tau}$ and $\emptyset \vdash \tau$.
+    \end{enumerate}
+\item If $\kappa = \kity$ then $\st=\sqity{\qity}$ and $\sofkn{\emptyset}{\emptyset}{\Phi}{0}{\st}{\kity}$ and $\svalNA{\st}$ and $\qtuq{\st'} \notin \qity$ and 
+    \begin{enumerate}
+    \item The outer form of $\qity$ is shared with $\tau$ and if $\qity'$ is a sub-term of $\qity$ then $\sofkn{\emptyset}{\emptyset}{\Phi}{0}{\sqity{\qity'}}{\kity}$ and $\svalNA{\sqity{\qity'}}$; or
+    \item $\qity = \srep{\st'}$ and $\istype{\st'}{\Phi}$
+    \end{enumerate}
+\item If $\kappa = \kitm$ then $\st=\sqitm{\qitm}$ and no sub-terms of $\qitm$ have the form $\quq{\st'}$ and 
+    \begin{enumerate}
+    \item The outer form of $\qitm$ is shared with $\itm$ and if $\qitm'$ is a sub-term of $\qitm$ then $\sofkn{\emptyset}{\emptyset}{\Phi}{n}{\sqitm{\qitm'}}{\kitm}$ and $\sval{\sqitm{\qitm'}}{\argEnv}$ and if $\qity$ is a sub-term of $\qitm$ then $\sofkn{\emptyset}{\emptyset}{\Phi}{0}{\sqity{\qity}}{\kity}$ and $\svalNA{\sqity{\qity}}$; or
+    \item $\qitm=\anatrans{n'}{\st'}$ and $n' < n$ and $\istype{\st'}{\Phi}$; or
+    \item $\qitm=\syntrans{n'}$ and $n' < n$.
+    \end{enumerate}
+\end{enumerate}
+\end{theorem}
+
+\begin{theorem}[Static Progress]
+If $\sofkn{\emptyset}{\emptyset}{\Phi}{n}{\st}{\kappa}$ and $\vdash \Phi$ and $\vdash^n \argEnv$ then $\sval{\st}{\argEnv}$ or $\serr{\st}{\argEnv}$ or $\sstep{\st}{\argEnv}{\st'}$.
+\end{theorem}
+\begin{proof}We proceed by rule induction on the kinding judgement.
+(k-parr)
+
+(k-ty)
+
+(k-otherty)
+
+(k-tycase-parr)
+
+(k-tycase)
+
+(k-ity-lam)
+
+(k-ity-alpha)
+
+(k-ity-unquote)
+
+(k-ity-trans)
+
+(k-raise)
+
+(k-itm-var)
+
+(k-itm-lam)
+
+(k-itm-unquote)
+
+(k-itm-anatrans)
+
+(k-itm-syntrans)
+\end{proof}
+
+\begin{theorem}[Static Preservation]
+If $\sofkn{\emptyset}{\emptyset}{\Phi}{n}{\st}{\kappa}$ and $\vdash \Phi$ and $\vdash^n \argEnv$ and $\sstep{\st}{\argEnv}{\st'}$ then $\sofkn{\emptyset}{\emptyset}{\Phi}{n}{\st'}{\kappa}$.
+\end{theorem}
+\begin{proof}
+(s-ty-step) (2 cases)
+(s-tycase-step) (2 cases)
+(s-tycase-match) (2 cases)
+(s-tycase-fail-1)
+(s-tycase-fail-2)
+(s-ity-lam-step-1)
+(s-ity-lam-step-2)
+(s-ity-unquote-step)
+(s-ity-unquote-elim)
+(s-ity-trans-step)
+(s-itm-lam-step-1)
+(s-itm-lam-step-2)
+(s-itm-unquote-step)
+(s-itm-unquote-elim)
+(s-ana-step)
+(s-ana-success)
+(s-syn-success) (relies on type synthesis theorem)
+(s-itm-anatrans-step)
+\end{proof}
+
+% \begin{lemma}[Kinding Stability]
+% If $\vdash \Phi$ and $\vdash \Phi, \tcdef{\tc}{\psi}{\theta}$ and $\kDelta \vdash \kGamma$ and $\sofkn{\kDelta}{\kGamma}{\Phi}{n}{\st}{\kappa}$ then $\sofkn{\kDelta}{\kGamma}{\Phi, \tcdef{\tc}{\psi}{\theta}}{n}{\st}{\kappa}$.
+% \end{lemma}
+% \begin{proof}
+% all cases straightforward by induction
+% \end{proof}
+
+% TODO: static normalization stability inside A?
+
+\subsection{Types}
+\begin{lemma}[Type Substitution Application]\label{thm:type-substitution-application}
+If $\vdash \delta : \Delta$ and $\Delta \vdash \tau$ then $\emptyset \vdash [\delta]\tau$.
+\end{lemma}
+
+\begin{lemma}[Selective Type Abstraction]\label{thm:selective-type-abstraction}
+If $\vdash \Phi$ and $\sofkz{\emptyset}{\emptyset}{\Phi}{\sqity{\qity}}{\kity}$ and $\svalNA{\sqity{\qity}}$ and $\memD \leadsto \delta : \Delta$ and $\vdash \delta : \Delta$ and $\tdeabs{\tc}{\Phi}{\qity}{\memD}{\tau}{\memD'}$ then $\memD' \leadsto \delta' : \Delta'$ and $\delta \subseteq \delta'$ and $\Delta \subseteq \Delta'$ and $\vdash \delta' : \Delta'$ and $\Delta' \vdash \tau$.
+\end{lemma}
+\begin{proof}
+(shared forms) (trans(st))
+\end{proof}
+
+\begin{lemma}[Type Translation]\label{type-translation}
+If $\vdash \Phi$ and $\istype{\st}{\Phi}$ and $\vdash_\Phi \st \leadsto \tau$ then $\emptyset \vdash \tau$.
+\end{lemma}
+\begin{proof} By Lemma \ref{thm:type-substitution-application} and Lemma \ref{thm:selective-type-abstraction}.
+\end{proof}
+
+\begin{lemma}[Typing Context Translation]
+If $\vdash \Phi$ and $\vdash_\Phi \Upsilon \leadsto \Gamma$ then $\emptyset \vdash \Gamma$.\end{lemma}
+\begin{proof} We proceed by rule induction on typing context translation.
+Empty case trivial.
+Extended case follows by Lemma \ref{type-translation} and definition of internal typing context formation.\end{proof}
+
+\subsection{External Terms}
+\begin{theorem}[Type Synthesis]
+If $\vdash \Phi$ and $\vdash_\Phi \Upsilon \leadsto \Gamma$ and $\esynX{e}{\st}{\iota}$ then $\istype{\st}{\Phi}$ and $\vdash_\Phi \st \leadsto \tau$. \end{theorem}
+\begin{proof}
+(var) (ap) (syn-targ)
+\end{proof}
+
+\begin{lemma}[Selective Term Abstraction]
+If $\vdash \Phi$ and $\vdash^n \argEnv$ and $\sofkn{\emptyset}{\emptyset}{\Phi}{n}{\sqitm{\qitm}}{\kitm}$ and $\sval{\sqitm{\qitm}}{\argEnv}$ and $\memD \leadsto \delta : \Delta$ and $\vdash \delta : \Delta$ and $\emptyset \vdash \Gamma_\text{out}$ and $\memG \leadsto \gamma : \Gamma$ and $\Delta~\Gamma_\text{out} \vdash \gamma : \Gamma$ and $\edeabs{\tc}{\argEnv}{\qitm}{\memD}{\memG}{\itm}{\memD'}{\memG'}$ then $\memD' \leadsto \delta' : \Delta'$ and $\memG' \leadsto \gamma' : \Gamma'$ and $\delta \subseteq \delta'$ and $\Delta \subseteq \Delta'$ and $\gamma \subseteq \gamma'$ and $\Gamma \subseteq \Gamma'$ and $\vdash \delta' : \Delta'$ and $\Delta'~\Gamma_\text{out} \vdash \gamma' : \Gamma'$.
+\end{lemma}
+\begin{proof} By rule induction on selective term abstraction judgement.
+
+(shared forms) (unquote form not possibly by canonical forms) (anatrans) (syntrans)
+\end{proof}
+
+\begin{theorem}[Type-Preserving Translation]
+If $\vdash \Phi$ and $\vdash_\Phi \Upsilon \leadsto \Gamma$ and either
+\begin{enumerate}
+\item $\eanaX{e}{\st}{\iota}$ and $\istype{\st}{\Phi}$; or 
+\item $\esynX{e}{\st}{\iota}$
+\end{enumerate}
+then $\vdash_\Phi \st \leadsto \tau$ then $\emptyset~\Gamma \vdash \iota : \tau$.
+\end{theorem}
+\begin{proof} By rule induction on typing rules.
+
+(var) (lam) (ap) (fix) (subsume) (ascribe) (ana-intro) (syn-targ) (other)
+\end{proof}
+
+TODO: stability of typing
+TODO: Unicity of typing
+TODO: hygiene?
+
+TODO: Make definitions out of these.
+The standard judgement $\Gamma \vdash \gamma : \Gamma'$ states that every binding $x : \tau$ in $\Gamma'$ has a corresponding substitution $\iota/x$ in $\gamma$ such that $\Gamma \vdash \iota : x$.
+
+, and  the judgement $\vdash \delta : \Delta$ checks that every type variable in $\Delta$ has a well-formed substitution in $\delta$
+
+\subsection{Conservativity}
+\begin{theorem}[Conservativity]
+If $\vdash \Phi$ and $\istype{\sty{\tc}{\sttyidx}}{\Phi}$ and for all $e$, if $\eana{\emptyset}{\Phi}{e}{\sty{\tc}{\sttyidx}}{\iota}$ and $\iota \Downarrow \iota'$ then $P(\iota')$, then if $\vdash \Phi, \tcdef{\tc'}{\psi}{\theta}$ then for all $e$, if $\eana{\emptyset}{\Phi, \tcdef{\tc'}{\psi}{\theta}}{e}{\sty{\tc}{\sttyidx}}{\iota}$ and $\iota \Downarrow \iota'$ then $P(\iota')$.
+\end{theorem}
+
+\begin{lemma}[Other Substitution]
+If $\vdash \Phi, \tcdef{\tc}{\psi}{\theta}$ and $\istype{\st}{\Phi}$ and $\eana{\Upsilon}{\Phi, \tcdef{\tc}{\psi}{\theta}}{e}{\st}{\iota}$ then there exists $e'$ such that $\eana{\Upsilon}{\Phi}{e'}{\st}{\iota}$.
+\end{lemma}
+
+\section{Related Work}\label{prior-work}
+
+\section{Conclusion}\label{conclusion}
 We combine several interesting type theoretic techniques, applying them to novel ends: 1)  a bidirectional type system permits flexible reuse of a fixed syntax; 2) the SL serves both as an extension language and as the type-level language; we give it its own statics (i.e.
@@
 Like modules, reasoning about these \emph{modular type constructors} does not require  mechanized specifications or proofs: correctness issues in the type constructor logic necessarily causes typechecking to fail, so even extensions that are not proven correct can be distributed and ``tested'' in the wild without compromising the integrity of an entire program (at worst, only values of types constructed by the tycon being tested may exhibit undesirable properties).
+
 % But mechanized proofs can be provided to verify that such failures (which do not indicate mistakes in client code but are more analagous to ``translation validation failures'' arising from the compiler \cite{Pnueli-Siegel-Singerman98}, will not occur during typechecking),
@@
 \ref{variants}).
-We conclude by discussing some limitations, proposing some richer variants of the calculus and discussing related work (Sec.
+%We conclude by discussing some limitations, proposing some richer variants of the calculus and discussing related work (Sec.
 \ref{prior-work}).
@@
 \section{Appendix}
+\subsection{More Examples}
 
+For example, the opcon definition for $\opname{prj}$ in $\tcvar{lprod}$ is:
+\[\small
+\begin{array}{lcl}
+\omega_\text{lprod/targops} & := & \keyw{syn}~\opname{prj}=\slam{\klist{\klbl\times\kty}}{\svar{tyidx}}{
+    \slam{\klbl}{\svar{tmidx}}{
+        \slam{\kargs}{\svar{args}}{\\&&
+            \quad \keyw{let}~\svar{self} :: \keyw{Arg} = \svar{arity1}~\svar{args}\\&&
+            \quad \keyw{letpair}~(\_, \svar{str}) = \svar{syn}~\svar{self}\\&&
+            \quad \keyw{letpair}~(\svar{oty}, \svar{tr})=\skap{\keyw{Opt}[\kty]\times\kitm}{\skap{\klbl\times\kty}{\svar{listrec}}}~\svar{tyidx}~(\skap{\kty}{\svar{none}}, \svar{str})\\&&
+            \quad\quad \slam{\klbl\times\kty}{\svar{row}}{
+                \slam{\keyw{Opt}[\kty]\times\kitm}{\svar{rout}}{\\&&
+                    \quad\quad\quad \keyw{letpair}~(\svar{rty}, \svar{rtr})=\svar{rout}\\&&
+                    \quad\quad\quad \skap{\keyw{Opt}[\kty]\times\kitm}{\skap{\kty}{\svar{optcase}}}~(\svar{rty})~(\slam{\kty}{\_}{\svar{rout}})~(\slam{\kunit}{\_}{\\&&
+                        \quad\quad\quad\quad\keyw{letpair}~(\svar{rowlbl}, \svar{rowty})=\svar{row}\\&&
+                        \quad\quad\quad\quad \svar{iflbleq}~\svar{rowlbl}~\svar{tmidx}~\\&&
+                        \quad\quad\quad\quad\quad (\slam{\kunit}{\_}{(\skap{\kty}{\svar{some}}~\svar{rowty}, \sqitm{\keyw{fst}(\quq{\svar{rtr}})})})\\&&
+                        \quad\quad\quad\quad\quad (\slam{\kunit}{\_}{(\svar{rty}, \sqitm{\keyw{snd}(\quq{\svar{rtr}})})}))
+                    }
+                }
+            }\\&&
+            \quad \svar{optcase}~\svar{oty}~(\slam{\kty}{\svar{ty}}{(\svar{ty}, \svar{tr})})~(\slam{\kunit}{\_}{\sraise{\kty\times\kitm}})
+        }
+    }
+},\\&&
+\omega_\text{lprod/targops/1}
+\end{array}
+\]
+The target term is passed in as the first argument.
+The first two lines simply check that no other arguments were passed in and (re-)synthesize the type and translation for the target.
+Then, we recurse over the type index looking for the provided label.
+We maintain a translation and a type option, which remains $\svar{none}$ until the label is found.
+The translation involves projecting out the second component of the translation of the labeled product until the label is found, at which point we project out the first component and stop.
+If this process succeeds, we return the resulting type and translation.
+Otherwise, we raise a type error.
+
+\begin{lemma}
+If $\sofkn{\kDelta}{\kGamma}{\Phi}{n'}{\st}{\kappa}$ and $n > n'
+$ then $\sofkn{\kDelta}{\kGamma}{\Phi}{n}{\st}{\kappa}$.
+\end{lemma}
+\begin{proof}
+We can proceed by straightforward rule induction, noting that all the kinding rules  have been defined for all $n$ except for the four most recently defined.
+In these cases, the transitive property of inequality on natural numbers implies that the same rule can directly be applied to derive the consequent in each  case.
+\end{proof}
+
+% \begin{figure}
+% \begin{mathpar}
+% \inferrule{a}{b}
+% \end{mathpar}
+% \caption{SL Dynamics}
+% \label{SL-dynamics}
+% \end{figure}
 \end{document}

### 2014-11-25 att-icfp14/att-icfp14.tex
Updated proposal

@@
 But this is not generally possible from within, so it is common to develop dialects and ``toy'' languages.
-Unfortunately, taking this approach limits the utility of these abstractions: they cannot be imported by clients of other dialects and languages, and building applications where different components rely on different type systems is both unsafe and unnatural.
+Unfortunately, taking this approach limits the utility of these abstractions: they cannot be imported by clients of other languages, and building applications where different components rely on different type systems is both unsafe and unnatural.
 %An {internally} {extensible} type system could help address these issues, but designing an extension mechanism that is both expressive and safe is non-trivial: extensions must not be permitted to compromise key metatheoretic properties of the language, nor should they be allowed to weaken one another, in any combination.
@@
 %Having established why a minimal language like $\mathcal{L}\{\rightarrow\forall\}$, while suitable as an internal language, needs to be extended with additional fragments before it is suitable for use as a human-facing external language,
-A strong embedding, i.e.
-an isomorphism that preserves both static reasoning principles and dynamic behaviors, in a sense that we will make more precise as we go on, is generally far more difficult to establish than a weak embedding.
+A stronger embedding requires constructing an isomorphism that preserves both static reasoning principles and dynamic behaviors, in a sense that we will make more precise as we go on.
+This is generally far more difficult to establish than a weak embedding, and fewer universality results can be applied.
 For example, the aforementioned embeddings into $\mathcal{L}\{\rightarrow\forall\}$ do not preserve type disequality and some other equational reasoning principles (we will see less subtle issues soon).
-Establishing a strong embedding that is natural (as measured, to a first approximation, by the ratio of boilerplate code that must be manually generated in the embedding vs.
+Establishing an isomorphic embedding that is natural (as measured, to a first approximation, by the ratio of boilerplate code that must be manually generated in the embedding vs.
 in the fragment, and how likely one is to make a mistake when writing it) and that has a reasonable cost semantics is harder still.
@@
 Modern languages like ML do, of course, provide constructs, like datatypes and abstract types, that admit strong and often satisfying embeddings of many language fragments, occupying what is widely seen as a ``sweet spot'' in the design space.
-However, ``general-purpose'' and ''all-purpose'' remain quite distinct, and situations continue to arise in both research and practice where desirable fragments can still only be weakly defined in terms of general-purpose constructs, or a strong embedding, while possible, is widely seen as too verbose or inefficient.
+However, ``general-purpose'' and ''all-purpose'' remain quite distinct, and situations continue to arise in both research and practice where desirable fragments can still only be weakly defined in terms of general-purpose constructs, or an isomorphic embedding, while possible, is widely seen as too verbose or inefficient.
 For example:
@@
 There are many  variants of product types: records, records with functional record update, labeled tuples (records that specify a field ordering) and record-like data structures with field delegation (of various sorts).
-These are  either awkward to strongly embed or cannot be.
+These are  either awkward to isomorphically embed or cannot be.
 Similarly, sum types also admit many variants (e.g.
@@
 Drawing from this approach, we aim to introduce a mechanism that makes it possible to declare new type and operator constructors and define, \emph{using type-level functions}, their static and dynamic semantics, the latter by translation to a fixed typed internal language.
-If a language fragment can be weakly defined in terms of the typed internal language and its statics are of a general form consistent with this protocol, a  strong embedding in the external language can be constructed by simply introducing the necessary constructors and defining directly the necessary type assignment and translation logic.
+If a language fragment can be weakly defined in terms of the typed internal language and its statics are of a general form consistent with this protocol, a n isomorphic embedding in the external language can be constructed by simply introducing the necessary constructors and defining directly the necessary type assignment and translation logic.
 We call such an embedding an \emph{active embedding}.
@@
 Natural numbers can, of course, be strongly embedded in existing languages, with a similar usage and asymptotic performance profile (up to function call overhead as an abstract type, for example).
-We will provide more sophisticated examples where this is less feasible later on (and note that type abstraction itself in Sec.
+We will provide more sophisticated examples where this is less feasible later on (and note that type abstraction itself .....
+in Sec.
 \ref{limitations}).\todo{if i don't have time/space, take this out} %It consists of:

### 2014-11-25 att-pldi15/att-pldi15-supplement.tex
Updated proposal

@@
 \section{Internal Language}
-\subsection{Semantics}
-We assume an internal language where the statics are specified in the standard way by judgements for  type formation {$\iDelta \vdash \tau$}, typing context formation { $\iDelta \vdash \iGamma$} and type assignment {$\iDelta~\iGamma \vdash \iota : \tau\moutput$}.
-%The typing and type formation contexts obey standard structural properties (i.e. weakening, exchange and contraction). 
-The internal dynamics are specified as a structural operational semantics with a stepping judgement {\small $\iota \mapsto \iota\moutput$} and a value judgement {$\iota~\mathtt{val}$}.
-The multi-step judgement $\iota \mapsto^{*} \iota\moutput$ is the reflexive, transitive closure of the stepping judgement and the evaluation judgement $\iota \Downarrow \iota'$ is defined iff $\iota \mapsto^{*} \iota'$ and $\iota'~\mathtt{val}$.
-Both the static and dynamic semantics of the IL can be found in any standard textbook covering typed lambda calculi (we directly follow \cite{pfpl}), so we assume familiarity and give the lemmas in this section without proof.
-
-Our particular choice in the paper is {$\mathcal{L}\{{\rightharpoonup}\,{\forall}\,{\mu}\,{\iunit}\,{\times}\,{+}\}$}, the syntax for which is shown in Figure \ref{syntax-IL}, as representative of any intermediate language for a typed functional language.
+Our particular choice of IL in the paper is {$\mathcal{L}\{{\rightharpoonup}\,{\forall}\,{\mu}\,{\iunit}\,{\times}\,{+}\}$}, the syntax for which is shown in Figure \ref{syntax-IL}, as representative of any intermediate language for a typed functional language.
 In fact, our intention is not to prescribe a particular choice of IL, so we will here only review the key metatheoretic properties that the IL must possess.
@@
 
-\begin{assumption}[Internal Type Assignment]
-If $\Delta \vdash \Gamma$ and $\Delta~\Gamma \vdash \iota : \tau$ then $\Delta \vdash \tau$.
-\end{assumption}
-
-Internal type safety follows the standard methodology: preservation and progress lemmas.
-
-\begin{assumption}[Internal Progress]
-If $\emptyset~\emptyset \vdash \iota : \tau$ then either $\iota~\texttt{val}$ or $\iota \mapsto \iota'$.
-\end{assumption}
-\begin{assumption}[Internal Preservation]
-If $\emptyset~\emptyset \vdash \iota : \tau$ and $\iota \mapsto \iota'$ then $\emptyset~\emptyset \vdash \iota' : \tau$. 
-\end{assumption}
+\subsection{Internal Statics}
+We assume an internal language where the statics are specified in the standard way by judgements for  type formation {$\iDelta \vdash \tau$}, shown in Figure \ref{fig:internal-type-formation}, typing context formation { $\iDelta \vdash \iGamma$}, shown in Figure \ref{fig:internal-typing-context-formation}, and type assignment {$\iDelta~\iGamma \vdash \iota : \tau\moutput$}, shown in Figure \ref{fig:internal-type-asignment}.
+\todo{do these and cite relevant rules in remainder}
+%The typing and type formation contexts obey standard structural properties (i.e. weakening, exchange and contraction).
+
+% The internal dynamics must be specified in a manner that leads to an appropriate type safety lemma (e.g.
+as a structural operational semantics with a stepping judgement {\small $\iota \mapsto \iota\moutput$} and a value judgement {$\iota~\mathtt{val}$}).
+The dynamic semantics of the IL can be found in any standard textbook covering typed lambda calculi (we directly follow \cite{pfpl}), so we assume familiarity.
+
+% \begin{assumption}[Internal Type Assignment]
+% If $\Delta \vdash \Gamma$ and $\Delta~\Gamma \vdash \iota : \tau$ then $\Delta \vdash \tau$.
+% \end{assumption}
+
+% Internal type safety follows the standard methodology: preservation and progress lemmas.
+
+% \begin{assumption}[Internal Progress]
+% If $\emptyset~\emptyset \vdash \iota : \tau$ then either $\iota~\texttt{val}$ or $\iota \mapsto \iota'$.
+% \end{assumption}
+% \begin{assumption}[Internal Preservation]
+% If $\emptyset~\emptyset \vdash \iota : \tau$ and $\iota \mapsto \iota'$ then $\emptyset~\emptyset \vdash \iota' : \tau$. 
+% \end{assumption}
 
@@
 
-\subsection{Explicit Substitutions}
+\subsection{Internal Substitutions}
 We also assume substitution is defined in the usual way.
@@
 
-\begin{assumption}[Internal Type Substitutions on Typing Contexts] 
-\label{lemma:internal-type-substitutions-on-typing-contexts}
-If $\Delta \vdash \delta : \Delta'$ and $\Delta\Delta' \vdash \Gamma$ then $\Delta \vdash [\delta]\Gamma$.
-\end{assumption}
+% \begin{assumption}[Internal Type Substitutions on Typing Contexts] 
+% \label{lemma:internal-type-substitutions-on-typing-contexts}
+% If $\Delta \vdash \delta : \Delta'$ and $\Delta\Delta' \vdash \Gamma$ then $\Delta \vdash [\delta]\Gamma$.
+% \end{assumption}
 
@@
 \subsubsection{Backprojection}
-To avoid having to give a large number of rules that do nothing interesting, we define backprojection on static terms definitionally.
-In the discussion below, let $\Phi'=\Phi, \tcdef{\tc'}{\tcsig{\ktyidx}{\chi}}{\tcstruct{\stx{schema}}{\omega}}}{\st}{\kappa}$.
-
-\newcommand{\bprj}[4]{{}^{m}\mathcal{B}_{#1}^{#2}(#3)=#4}
-\newcommand{\bprjX}[2]{\bprj{\Phi'}{\tc'}{#1}{#2}}
+To avoid having to give a large number of rules that do nothing interesting, we define backprojection on static terms definitionally.\todo{eval of index should be empty Phi}
+
 \begin{definition}[Backprojection (Static values)]

### 2014-11-25 ecoop12/clq-ecoop12.tex
Updated proposal

--- 2012-08-30 ipdps13/clq-ipdps13.tex
@@
-%% bare_conf.tex
-%% V1.3
-%% 2007/01/11
-%% by Michael Shell
-%% See:
-%% http://www.michaelshell.org/
-%% for current contact information.
-%%
-%% This is a skeleton file demonstrating the use of IEEEtran.cls
-%% (requires IEEEtran.cls version 1.7 or later) with an IEEE conference paper.
-%%
-%% Support sites:
-%% http://www.michaelshell.org/tex/ieeetran/
-%% http://www.ctan.org/tex-archive/macros/latex/contrib/IEEEtran/
-%% and
-%% http://www.ieee.org/
-
-%%*************************************************************************
-%% Legal Notice:
-%% This code is offered as-is without any warranty either expressed or
-%% implied; without even the implied warranty of MERCHANTABILITY or
-%% FITNESS FOR A PARTICULAR PURPOSE! 
-%% User assumes all risk.
-%% In no event shall IEEE or any contributor to this code be liable for
-%% any damages or losses, including, but not limited to, incidental,
-%% consequential, or any other damages, resulting from the use or misuse
-%% of any information contained here.
-%%
-%% All comments are the opinions of their respective authors and are not
-%% necessarily endorsed by the IEEE.
-%%
-%% This work is distributed under the LaTeX Project Public License (LPPL)
-%% ( http://www.latex-project.org/ ) version 1.3, and may be freely used,
-%% distributed and modified. A copy of the LPPL, version 1.3, is included
-%% in the base LaTeX documentation of all distributions of LaTeX released
-%% 2003/12/01 or later.
-%% Retain all contribution notices and credits.
-%% ** Modified files should be clearly indicated as such, including  **
-%% ** renaming them and changing author support contact information. **
-%%
-%% File list of work: IEEEtran.cls, IEEEtran_HOWTO.pdf, bare_adv.tex,
-%%                    bare_conf.tex, bare_jrnl.tex, bare_jrnl_compsoc.tex
-%%*************************************************************************
-
-% *** Authors should verify (and, if needed, correct) their LaTeX system  ***
-% *** with the testflow diagnostic prior to trusting their LaTeX platform ***
-% *** with production work. IEEE's font choices can trigger bugs that do  ***
-% *** not appear when using other class files.                            ***
-% The testflow support page is at:
-% http://www.michaelshell.org/tex/testflow/
-
-% Note that the a4paper option is mainly intended so that authors in
-% countries using A4 can easily print to A4 and see how their papers will
-% look in print - the typesetting of the document will not typically be
-% affected with changes in paper size (but the bottom and side margins will).
-% Use the testflow package mentioned above to verify correct handling of
-% both paper sizes by the user's LaTeX system.
+% This is LLNCS.DEM the demonstration file of
+% the LaTeX macro package from Springer-Verlag
+% for Lecture Notes in Computer Science,
+% version 2.2 for LaTeX2e
 %
-% Also note that the "draftcls" or "draftclsnofoot", not "draft", option
-% should be used if it is desired that the figures are to be displayed in
-% draft mode.
+\documentclass{llncs}
 %
-\documentclass[10pt, conference, compsocconf]{IEEEtran}
-% Add the compsocconf option for Computer Society conferences.
+\usepackage{makeidx}  % allows for indexgeneration
 %
-% If IEEEtran.cls has not been installed into the LaTeX system files,
-% manually specify the path to it like:
-% \documentclass[conference]{../sty/IEEEtran}
-
-% Some very useful LaTeX packages include:
-% (uncomment the ones you want to load)
-
-% *** MISC UTILITY PACKAGES ***
-%
-%\usepackage{ifpdf}
-% Heiko Oberdiek's ifpdf.sty is very useful if you need conditional
-% compilation based on whether the output is pdf or dvi.
-% usage:
-% \ifpdf
-%   % pdf code
-% \else
-%   % dvi code
-% \fi
-% The latest version of ifpdf.sty can be obtained from:
-% http://www.ctan.org/tex-archive/macros/latex/contrib/oberdiek/
-% Also, note that IEEEtran.cls V1.7 and later provides a builtin
-% \ifCLASSINFOpdf conditional that works the same way.
-% When switching from latex to pdflatex and vice-versa, the compiler may
-% have to be run twice to clear warning/error messages.
-
-% *** CITATION PACKAGES ***
-%
-%\usepackage{cite}
-% cite.sty was written by Donald Arseneau
-% V1.6 and later of IEEEtran pre-defines the format of the cite.sty package
-% \cite{} output to follow that of IEEE. Loading the cite package will
-% result in citation numbers being automatically sorted and properly
-% "compressed/ranged". e.g., [1], [9], [2], [7], [5], [6] without using
-% cite.sty will become [1], [2], [5]--[7], [9] using cite.sty. cite.sty's
-% \cite will automatically add leading space, if needed. Use cite.sty's
-% noadjust option (cite.sty V3.8 and later) if you want to turn this off.
-% cite.sty is already installed on most LaTeX systems. Be sure and use
-% version 4.0 (2003-05-27) and later if using hyperref.sty. cite.sty does
-% not currently provide for hyperlinked citations.
-% The latest version can be obtained at:
-% http://www.ctan.org/tex-archive/macros/latex/contrib/cite/
-% The documentation is contained in the cite.sty file itself.
-
-% *** GRAPHICS RELATED PACKAGES ***
-%
-\ifCLASSINFOpdf
-  % \usepackage[pdftex]{graphicx}
-  % declare the path(s) where your graphic files are
-  % \graphicspath{{../pdf/}{../jpeg/}}
-  % and their extensions so you won't have to specify these with
-  % every instance of \includegraphics
-  % \DeclareGraphicsExtensions{.pdf,.jpeg,.png}
-\else
-  % or other class option (dvipsone, dvipdf, if not using dvips). graphicx
-  % will default to the driver specified in the system graphics.cfg if no
-  % driver is specified.
-  % \usepackage[dvips]{graphicx}
-  % declare the path(s) where your graphic files are
-  % \graphicspath{{../eps/}}
-  % and their extensions so you won't have to specify these with
-  % every instance of \includegraphics
-  % \DeclareGraphicsExtensions{.eps}
-\fi
-% graphicx was written by David Carlisle and Sebastian Rahtz. It is
-% required if you want graphics, photos, etc. graphicx.sty is already
-% installed on most LaTeX systems. The latest version and documentation can
-% be obtained at: 
-% http://www.ctan.org/tex-archive/macros/latex/required/graphics/
-% Another good source of documentation is "Using Imported Graphics in
-% LaTeX2e" by Keith Reckdahl which can be found as epslatex.ps or
-% epslatex.pdf at: http://www.ctan.org/tex-archive/info/
-%
-% latex, and pdflatex in dvi mode, support graphics in encapsulated
-% postscript (.eps) format. pdflatex in pdf mode supports graphics
-% in .pdf, .jpeg, .png and .mps (metapost) formats. Users should ensure
-% that all non-photo figures use a vector format (.eps, .pdf, .mps) and
-% not a bitmapped formats (.jpeg, .png). IEEE frowns on bitmapped formats
-% which can result in "jaggedy"/blurry rendering of lines and letters as
-% well as large increases in file sizes.
-%
-% You can find documentation about the pdfTeX application at:
-% http://www.tug.org/applications/pdftex
-
-% *** MATH PACKAGES ***
-%
-%\usepackage[cmex10]{amsmath}
-% A popular package from the American Mathematical Society that provides
-% many useful and powerful commands for dealing with mathematics. If using
-% it, be sure to load this package with the cmex10 option to ensure that
-% only type 1 fonts will utilized at all point sizes. Without this option,
-% it is possible that some math symbols, particularly those within
-% footnotes, will be rendered in bitmap form which will result in a
-% document that can not be IEEE Xplore compliant!
-%
-% Also, note that the amsmath package sets \interdisplaylinepenalty to 10000
-% thus preventing page breaks from occurring within multiline equations. Use:
-%\interdisplaylinepenalty=2500
-% after loading amsmath to restore such page breaks as IEEEtran.cls normally
-% does. amsmath.sty is already installed on most LaTeX systems. The latest
-% version and documentation can be obtained at:
-% http://www.ctan.org/tex-archive/macros/latex/required/amslatex/math/
-
-% *** SPECIALIZED LIST PACKAGES ***
-%
-%\usepackage{algorithmic}
-% algorithmic.sty was written by Peter Williams and Rogerio Brito.
-% This package provides an algorithmic environment fo describing algorithms.
-% You can use the algorithmic environment in-text or within a figure
-% environment to provide for a floating algorithm. Do NOT use the algorithm
-% floating environment provided by algorithm.sty (by the same authors) or
-% algorithm2e.sty (by Christophe Fiorio) as IEEE does not use dedicated
-% algorithm float types and packages that provide these will not provide
-% correct IEEE style captions. The latest version and documentation of
-% algorithmic.sty can be obtained at:
-% http://www.ctan.org/tex-archive/macros/latex/contrib/algorithms/
-% There is also a support site at:
-% http://algorithms.berlios.de/index.html
-% Also of interest may be the (relatively newer and more customizable)
-% algorithmicx.sty package by Szasz Janos:
-% http://www.ctan.org/tex-archive/macros/latex/contrib/algorithmicx/
-
-% *** ALIGNMENT PACKAGES ***
-%
-%\usepackage{array}
-% Frank Mittelbach's and David Carlisle's array.sty patches and improves
-% the standard LaTeX2e array and tabular environments to provide better
-% appearance and additional user controls. As the default LaTeX2e table
-% generation code is lacking to the point of almost being broken with
-% respect to the quality of the end results, all users are strongly
-% advised to use an enhanced (at the very least that provided by array.sty)
-% set of table tools. array.sty is already installed on most systems. The
-% latest version and documentation can be obtained at:
-% http://www.ctan.org/tex-archive/macros/latex/required/tools/
-
-%\usepackage{mdwmath}
-%\usepackage{mdwtab}
-% Also highly recommended is Mark Wooding's extremely powerful MDW tools,
-% especially mdwmath.sty and mdwtab.sty which are used to format equations
-% and tables, respectively. The MDWtools set is already installed on most
-% LaTeX systems. The lastest version and documentation is available at:
-% http://www.ctan.org/tex-archive/macros/latex/contrib/mdwtools/
-
-% IEEEtran contains the IEEEeqnarray family of commands that can be used to
-% generate multiline equations as well as matrices, tables, etc., of high
-% quality.
-
-%\usepackage{eqparbox}
-% Also of notable interest is Scott Pakin's eqparbox package for creating
-% (automatically sized) equal width boxes - aka "natural width parboxes".
-% Available at:
-% http://www.ctan.org/tex-archive/macros/latex/contrib/eqparbox/
-
-% *** SUBFIGURE PACKAGES ***
-%\usepackage[tight,footnotesize]{subfigure}
-% subfigure.sty was written by Steven Douglas Cochran. This package makes it
-% easy to put subfigures in your figures. e.g., "Figure 1a and 1b". For IEEE
-% work, it is a good idea to load it with the tight package option to reduce
-% the amount of white space around the subfigures. subfigure.sty is already
-% installed on most LaTeX systems. The latest version and documentation can
-% be obtained at:
-% http://www.ctan.org/tex-archive/obsolete/macros/latex/contrib/subfigure/
-% subfigure.sty has been superceeded by subfig.sty.
-
-%\usepackage[caption=false]{caption}
-%\usepackage[font=footnotesize]{subfig}
-% subfig.sty, also written by Steven Douglas Cochran, is the modern
-% replacement for subfigure.sty. However, subfig.sty requires and
-% automatically loads Axel Sommerfeldt's caption.sty which will override
-% IEEEtran.cls handling of captions and this will result in nonIEEE style
-% figure/table captions. To prevent this problem, be sure and preload
-% caption.sty with its "caption=false" package option. This is will preserve
-% IEEEtran.cls handing of captions. Version 1.3 (2005/06/28) and later 
-% (recommended due to many improvements over 1.2) of subfig.sty supports
-% the caption=false option directly:
-%\usepackage[caption=false,font=footnotesize]{subfig}
-%
-% The latest version and documentation can be obtained at:
-% http://www.ctan.org/tex-archive/macros/latex/contrib/subfig/
-% The latest version and documentation of caption.sty can be obtained at:
-% http://www.ctan.org/tex-archive/macros/latex/contrib/caption/
-
-% *** FLOAT PACKAGES ***
-%
-%\usepackage{fixltx2e}
-% fixltx2e, the successor to the earlier fix2col.sty, was written by
-% Frank Mittelbach and David Carlisle. This package corrects a few problems
-% in the LaTeX2e kernel, the most notable of which is that in current
-% LaTeX2e releases, the ordering of single and double column floats is not
-% guaranteed to be preserved. Thus, an unpatched LaTeX2e can allow a
-% single column figure to be placed prior to an earlier double column
-% figure. The latest version and documentation can be found at:
-% http://www.ctan.org/tex-archive/macros/latex/base/
-
-%\usepackage{stfloats}
-% stfloats.sty was written by Sigitas Tolusis. This package gives LaTeX2e
-% the ability to do double column floats at the bottom of the page as well
-% as the top. (e.g., "\begin{figure*}[!b]" is not normally possible in
-% LaTeX2e). It also provides a command:
-%\fnbelowfloat
-% to enable the placement of footnotes below bottom floats (the standard
-% LaTeX2e kernel puts them above bottom floats). This is an invasive package
-% which rewrites many portions of the LaTeX2e float routines. It may not work
-% with other packages that modify the LaTeX2e float routines. The latest
-% version and documentation can be obtained at:
-% http://www.ctan.org/tex-archive/macros/latex/contrib/sttools/
-% Documentation is contained in the stfloats.sty comments as well as in the
-% presfull.pdf file. Do not use the stfloats baselinefloat ability as IEEE
-% does not allow \baselineskip to stretch. Authors submitting work to the
-% IEEE should note that IEEE rarely uses double column equations and
-% that authors should try to avoid such use. Do not be tempted to use the
-% cuted.sty or midfloat.sty packages (also by Sigitas Tolusis) as IEEE does
-% not format its papers in such ways.
-
-% *** PDF, URL AND HYPERLINK PACKAGES ***
-%
-%\usepackage{url}
-% url.sty was written by Donald Arseneau. It provides better support for
-% handling and breaking URLs. url.sty is already installed on most LaTeX
-% systems. The latest version can be obtained at:
-% http://www.ctan.org/tex-archive/macros/latex/contrib/misc/
-% Read the url.sty source comments for usage information. Basically,
-% \url{my_url_here}.
-
-% *** Do not adjust lengths that control margins, column widths, etc. ***
-% *** Do not use packages that alter fonts (such as pslatex).         ***
-% There should be no need to do such things with IEEEtran.cls V1.6 and later.
-% (Unless specifically asked to do so by the journal or conference you plan
-% to submit to, of course. )
-
-% correct bad hyphenation here
-\hyphenation{op-tical net-works semi-conduc-tor}
 \input{macros}
-
 \begin{document}
 %
-% paper title
-% can use linebreaks \\ within to get better formatting as desired
-\title{\Ace: An Actively-Typed Compilation Environment\\for High-Performance Computing}
+\frontmatter          % for the preliminaries
+%
+\pagestyle{headings}  % switches on printing of running heads
+\addtocmark{Hamiltonian Mechanics} % additional mark in the TOC
+%
+\mainmatter              % start of the contributions
+%
+\title{Practical Language Foundations for Scientific and High-Performance Computing}
+%
+\titlerunning{Language Foundations for HPC}  % abbreviated title (for running head)
+%                                     also used for the TOC unless
+%                                     \toctitle is used
+%
+\author{Cyrus Omar, Jonathan Aldrich}
+%
+\authorrunning{Cyrus Omar, Jonathan Aldrich}   % abbreviated author list (for running head)
+%
+%%%% modified list of authors for the TOC (add the affiliations)
+\tocauthor{Ivar Ekeland (Princeton University),
+Roger Temam (Universit\'{e} de Paris-Sud),
+Jeffrey Dean, David Grove, Craig Chambers (Universit\`a di Geova),
+Kim B. Bruce (Stanford University),
+Elisa Bertino (Digita Research Center)}
+%
+\institute{Carnegie Mellon University, Pittsburgh, PA, USA,\\
+\email{\{comar,jonathan.aldrich\}@cs.cmu.edu},\\\texttt{http://github.com/cyrus-/cl.oquence/}}
 
-% author names and affiliations
-% use a multiple column layout for up to two different
-% affiliations
-
-\author{\IEEEauthorblockN{Cyrus Omar, Jonathan Aldrich}
-\IEEEauthorblockA{School of Computer Science\\
-Carnegie Mellon University\\
-Pittsburgh, PA, USA\\
-\texttt{\{comar,jonathan.aldrich\}@cs.cmu.edu}}}
-
-% conference papers do not typically use \thanks and this command
-% is locked out in conference mode. If really needed, such as for
-% the acknowledgment of grants, issue a \IEEEoverridecommandlockouts
-% after \documentclass
-
-% for over three affiliations, or if they all won't fit within the width
-% of the page, use this alternative format:
-% 
-%\author{\IEEEauthorblockN{Michael Shell\IEEEauthorrefmark{1},
-%Homer Simpson\IEEEauthorrefmark{2},
-%James Kirk\IEEEauthorrefmark{3}, 
-%Montgomery Scott\IEEEauthorrefmark{3} and
-%Eldon Tyrell\IEEEauthorrefmark{4}}
-%\IEEEauthorblockA{\IEEEauthorrefmark{1}School of Electrical and Computer Engineering\\
-%Georgia Institute of Technology,
-%Atlanta, Georgia 30332--0250\\ Email: see http://www.michaelshell.org/contact.html}
-%\IEEEauthorblockA{\IEEEauthorrefmark{2}Twentieth Century Fox, Springfield, USA\\
-%Email: homer@thesimpsons.com}
-%\IEEEauthorblockA{\IEEEauthorrefmark{3}Starfleet Academy, San Francisco, California 96678-2391\\
-%Telephone: (800) 555--1212, Fax: (888) 555--1212}
-%\IEEEauthorblockA{\IEEEauthorrefmark{4}Tyrell Inc., 123 Replicant Street, Los Angeles, California 90210--4321}}
-
-% use for special paper notices
-%\IEEEspecialpapernotice{(Invited Paper)}
-
-% make the title area
-\maketitle
+\maketitle              % typeset the title of the contribution
 
 \begin{abstract}
-In this paper, we introduce the Ace programming language and demonstrate its practicality as a foundational tool for both high-performance computing researchers and practitioners.
-Ace is a statically-typed language that shares a syntax with Python and uses the Python language itself as a multi-purpose compile-time metalanguage.
-Ace has a minimal core that delegates control over many aspects of compilation to user modules via a novel extension mechanism called {\em active type-checking and translation (\ATT)}.
-Specifically, users specify new primitive types and operations by equipping type definitions -- first-class objects in the metalanguage -- with compile-time methods.
-The compiler invokes these methods when type checking and translating expressions, according to a fixed type dispatch protocol associated with each syntactic form.
+As scientific applications and high-performance computing architectures\- become increasingly complex, there is a growing need for languages and abstractions that achieve a better balance between raw performance,  productivity, portability and verifiability.
+Unfortunately, promising new approaches are adopted only slowly in practice.
+To understand why, we begin by reviewing several empirical studies of software development practices in  science and engineering.
+These studies suggest a number of non-trivial design and adoption criteria that constrain such efforts.
+With these criteria in mind, we design a new language: $\cloquence$.
+By employing a novel collection of front-end techniques and building upon established back-end infrastructure, $\cloquence$ satisfies  a number of seemingly conflicting criteria related to simplicity, expressive\- power, performance, portability and compatibility with existing code bases and tools.
+$\cloquence$ features a powerful type-based extension mechanism, suggesting a foundational role for the language in enabling the development and deployment of higher-level abstractions and new language features as {\it active libraries}.
+We demonstrate this flexibility by describing how to implement several common parallel abstractions that are today most cleanly implemented in their own standalone languages, focusing primarily on the OpenCL language, used to write low-level data-parallel programs for many-core processors, particularly GPUs.
+We conclude with a brief case study where $\cloquence$ was used to improve the structure of code written to simulate large-scale spiking neural circuits using GPUs.
+\end{abstract}
+%
+\section{Introduction}
+Computer-aided simulations and data analysis techniques have transformed science and engineering.
+Surveys show that scientists and engineers now spend up to 40\% of their time writing software, typically individually or in small groups \cite{howison2011scientific}\cite{hannay2009scientists}.
+Most of this software targets conventional desktop hardware, but about 20\% of scientists also target either local clusters or super\-computers for more numerically-intensive computations \cite{hannay2009scientists}.
+In recent years, GPUs and other kinds of coprocessors have emerged as powerful platforms for high-performance computing on both desktops and clusters as well.
 
-\end{abstract}
+Scientists and engineers generally prefer high-level scripting languages like MATLAB, Python, R and Perl \cite{nguyen2010survey}, particularly those bundled with powerful domain-specific libraries.
+These languages are typically interpreted rather than compiled and generally lack static type systems, relying instead on run-time (or ``dynamic'') type lookup and error handling.
+Although this may increase the flexibility of the language, these features also negatively impact performance and can lead to unexpected run-time errors that are generally impossible to rule out statically.
+This is particularly severe given that rigorous software testing practices are also quite rare in these domains \cite{nguyen2010survey}.
 
-%\begin{IEEEkeywords}
-%programming languages, heterogeneous computing, metaprogramming, code generation
+Along code paths where performance appears to be a bottleneck, developers will turn to statically-typed, though still unsafe, low-level languages like C and Fortran \cite{4222616}.
+These languages give users explicit control over data layout and heap allocation and require explicit type annotations on all variables.
+Although this makes writing programs more tedious which can increase the number of logic errors a programmer makes, it can also potentially improve performance, particularly if the programmer understands the target architecture well (e.g.
+cache behavior).
+While a sequential reimplementation of a critical code path using a low-level language will generally produce a modest speedup, more significant speedups on modern hardware require parallelizing over many processor cores and, on massively-parallel machines, many interconnected nodes.
+Despite well-known difficulties, low-level, unsafe  approaches that use a form of shared-memory  multithreading (e.g.
+pthreads, OpenMP) paired with explicit message passing between nodes (typically using MPI) remain widely used for these tasks as well \cite{4222616}\cite{basili2008understanding}.
 
-%\end{IEEEkeywords}
-
-% For peer review papers, you can put extra information on the cover
-% page as needed:
-% \ifCLASSOPTIONpeerreview
-% \begin{center} \bfseries EDICS Category: 3-BBND \end{center}
-% \fi
-%
-% For peerreview papers, this IEEEtran command inserts a page break and
-% creates the second title. It will be ignored for other modes.
-\IEEEpeerreviewmaketitle
-
-\section{Introduction}
-Researchers design and implement novel parallel programming abstractions\- on a regular basis.
-Each abstraction aims to make some class of developers more productive by making appropriate trade-offs between performance, portability, verifiability and ease-of-use.
-It can be observed that the most widely-adopted abstractions to date are those that been implemented as libraries that can be easily deployed within existing, widely-used languages.
-Abstractions that require introducing new primitive constructs, changing the semantics of existing constructs in particular contexts, or controlling compilation in a fine-grained manner cannot take this approach.
-This has led to a proliferation of specialized programming languages, each designed around a few privileged abstractions.
-Unfortunately, such languages have not been widely adopted by practitioners.
-
-Computer-aided simulations and data analysis techniques have transformed science and engineering.
-Surveys show that scientists and engineers now spend up to 40\% of their time writing software, generally individually or in small groups \cite{howison2011scientific}\cite{hannay2009scientists}.
-Most of this software targets conventional desktop hardware, but about 20\% of scientists also target either local clusters or super\-computers for more numerically-intensive computations \cite{hannay2009scientists}.
-In recent years, GPUs and other kinds of coprocessors have emerged as powerful platforms for high-performance computing as well.
-
-Many scientists and engineers prefer high-level languages like MATLAB, Python, R and Perl \cite{nguyen2010survey}, due to their powerful domain-specific libraries and their focus on productivity and flexibility.
-These languages are typically interpreted rather than compiled and generally do not feature static type systems, relying instead on run-time (``dynamic'') type lookup and indirection.
-Although this can increase the flexibility of the language, these features also negatively impact  performance.
-Along code paths where performance appears to be a bottleneck, developers typically turn to traditional statically-typed low-level languages like C and Fortran \cite{4222616}.
-These languages require explicit type annotations on variables and give users explicit control over data layout and heap allocation.
-Although this makes writing programs more tedious and error-prone, it can also significantly  improve performance, particularly if attributes of the target architecture (e.g.
-caching behavior) are considered.
-
-While a sequential reimplementation of a critical code path using a low-level language will produce a modest speedup, more significant speedups on modern hardware require parallelizing over many processor cores and, on massively-parallel machines, many interconnected nodes.
-Despite well-known difficulties, low-level approaches that use a form of shared-memory  multithreading (e.g.
-pthreads, OpenMP) paired with explicit message passing between nodes (typically using MPI) remain widely used \cite{4222616}\cite{basili2008understanding}.
-
-Although researchers often propose higher-level language features and parallel programming abstractions that aim to strike a better balance between raw performance, productivity, code portability and verifiability (see Section 2), end-users remain skeptical of new approaches.
-This viewpoint was perhaps most succinctly expressed by a participant interviewed in a recent study \cite{basili2008understanding}, who stated ``I hate MPI, I hate C++.
+Although researchers often propose higher-level language features and new parallel programming abstractions that aim to strike a better balance between raw performance, productivity, code portability and verifiability (see Section 2), end-users remain skeptical of new approaches.
+This viewpoint was perhaps most succinctly expressed by a participant interviewed in a recent study by Basili et al.
+\cite{basili2008understanding}, who stated ``I hate MPI, I hate C++.
 [But] if I had to choose again, I would probably choose the same.'' Although this sentiment is easy to dismiss as paradoxical, we believe that it demands direct examination by those in the research community working to advance the practice of scientific and high-performance computing.
-We begin in Section 2 by developing a set of design and adoption criteria for new languages and abstractions based on prior empirical studies of these {\it professional end-user developers} as well as our observations of characteristics common to successful projects in the past.
+We begin in Section 2 by developing a set of design and adoption criteria for new languages and abstractions based on prior empirical studies of this class of {\it professional end-user developer}, as well as our observations of characteristics common to successful projects in the past.
 
@@
 
-We then introduce a new programming language named $\cloquence$ (pronounced ``C eloquence'') in Section 3.
-$\cloquence$ initially targets code paths where low-level languages, particularly OpenCL, would be used today.
-By employing a collection of powerful front-end language and compiler design techniques, making pragmatic design choices, and building on top of established infrastructure, we argue that $\cloquence$ may uniquely satisfy many of the criteria established in Section 2, particularly those related to ease of use, familiarity, performance and tool support.
-
-Unlike many existing languages, $\cloquence$ is designed to be highly extensible from inception, employing a powerful compile-time language extension and code generation mechanism based around the concept of {\it active libraries} \cite{activelibraries}.
-We argue that this feature may make $\cloquence$ particularly suitable as a platform for researchers developing new parallel abstractions and other specialized, high-level language features, as well as  development tools.
+We then turn in Section 3 to the main purpose of this paper, which is to introduce a new programming language named $\cloquence$ (pronounced ``C eloquence'').
+$\cloquence$ initially targets code paths where low-level languages, particularly C and OpenCL, would be used today.
+By employing a collection of powerful front-end language and compiler design techniques, making pragmatic design choices, and building on top of established infrastructure, we argue that $\cloquence$ may satisfy many of the criteria established in Section 2, particularly those related to ease of use, familiarity, performance and tool support, in ways that previous efforts have not.
+Unlike many existing languages, $\cloquence$ is designed to be internally extensible, employing a powerful compile-time language extension and code generation mechanism based around a concept called {\it active libraries} \cite{activelibraries}.
+We argue that this feature may make $\cloquence$ particularly suitable as a platform for researchers developing new parallel abstractions and other specialized, high-level language features, as well as  other kinds of development tools and software frameworks.
 
 In Section 4 we briefly describe a case study where the language was used to conduct large-scale spiking neural circuit simulations on GPU hardware, demonstrating the feasibility of this approach on a non-trivial problem.
-Finally, conclude in Section 6 with future directions for both the $\cloquence$ language and the research community.
+Finally, we conclude in Section 5 with future directions for both the $\cloquence$ language and the research community more broadly.
 
@@
 Unfortunately, as in many areas of design, it can be difficult to objectively evaluate the merit of such efforts.
-To better support such evaluations, researchers in design disciplines typically develop a set of high-level  design criteria that serve as a rubric for evaluating and guiding their design efforts.
+To better support such evaluations, researchers in design disciplines typically develop a set of high-level  design criteria that serve as a rubric for evaluating and guiding their ongoing design efforts.
 In programming language design, particularly for scientific and high-performance computing, there have been few concerted efforts to develop a coherent set of design criteria that capture the needs of the targeted developer communities.
 Similarly, there have been few treatments of adoption criteria for new languages and abstractions in practice, an important issue given the slow rates of adoption today.
-We have purposefully separated these criteria from the specific language we describe in Section 3 in the hopes of starting a productive discussion within the community.
+For these reasons, we purposefully separate these criteria from the specific language we describe in Section 3.
+By doing so, we hope these criteria will be broadly useful to the research community and highlight our user-focused language design methodology.
 
 \subsection{Professional End-User Developers}
-An important first step is to identify the developer communities targeted by our design efforts.
+An important first step is to identify the developer communities targeted by a design effort.
 Researchers in software engineering typically distinguish between {\it end-user developers} and {\it professional developers}.
@@
 
-However, one can also go too far, so these benefits must be balanced with concerns about readability and familiarity.
+However, the benefits of concise syntax must be balanced with concerns about readability and familiarity.
 A number of principles have been proposed to operationalize the notion of code readability.
@@
 Most researchers become familiar with formal notation by studying mathematics.
-In nearly all commonly used languages in scientific computing, the common mathematical notation is used whenever possible.
+In nearly all commonly used languages in scientific computing, common mathematical notation is generally preferred.
 In contrast, many academic languages have settled on alternative notational styles.
@@
 Additionally, a few languages, such as Mathematica, have support for more typographically-rich mathematical notation via a structural editing interface.
-These techniques appear to be helpful, although we do not know of formal studies that provide evidence for this claim.
+This technique appears to be helpful, although we do not know of formal studies that provide evidence for this claim.
 
@@
 Modern software development now relies on a number of tools to assist with common tasks, such as debuggers,  profilers, syntax-aware editors, documentation generators, style checkers and interactive interpreters.
-Similarly, most established languages benefit from a centralized packaged repository (e.g.
-PyPI) and a package manager that can automate the installation of new packages.
-Such support is not trivial to develop from scratch, but few early adopters are likely willing to do without these tools for non-trivial projects \cite{squires2005programmers}.
+Similarly, most established languages benefit from a centralized package repository (e.g.
+PyPI) paired with a package manager that can automate the installation of new packages.
+These tools and infrastructure are not trivial to develop from scratch, and few incentives exist for researchers to do so, but few users are willing to do without these tools for non-trivial projects \cite{squires2005programmers}.
 
@@
 \section{The $\cloquence$ Language}
-We now describe a programming language designed with many of the criteria described in the previous sections in mind, $\cloquence$.
+We now describe a programming language designed with many of the criteria described in the previous sections in mind.
 $\cloquence$ is built around a minimal core and nearly every construct in the language is implemented as a library.
@@
 Each thread, indexed by \verb|gid|, applies the transfer function, \verb|fn|, to the  corresponding element of the input array, \verb|in|, writing the result into the corresponding location in the output array, \verb|out|.
-
-\subsection{Syntax}
-Readers familiar with the Python programming language will recognize the style of syntax used in Figure \ref{map}.
-In fact, $\cloquence$ uses the Python grammar and parsing facilities directly.
-Several factors motivated this design decision.
-First, Python's syntax is widely credited as being particularly simple and readable, due to its use of significant whitespace and conventional mathematical notation.
-Python is one of the most widely-used languages in scientific computing, so its syntax is already familiar to much of the field.
-And significantly, a large ecosystem of tools already exist that work with Python files, such as code editors, syntax highlighters, style checkers and documentation generators.
-These can be used without modification to work with $\cloquence$ files.
-Therefore, by re-using an existing, widely-used grammar, we are able to satisfy many of the design criteria described in Section \ref{syntax} and the adoption criteria described in Section \ref{tools} without significant development effort.
 
@@
 
+\subsection{Syntax}
+Readers familiar with the Python programming language will recognize the style of syntax used in Figure \ref{map}.
+In fact, $\cloquence$ uses the Python grammar and parsing facilities directly.
+Several factors motivated this design decision.
+First, Python's syntax is widely credited as being particularly simple and readable, due to its use of significant whitespace and conventional mathematical notation.
+Python is one of the most widely-used languages in scientific computing, so its syntax is already familiar to much of the field.
+And significantly, a large ecosystem of tools already exist that work with Python files, such as code editors, syntax highlighters, style checkers and documentation generators.
+These can be used without modification to work with $\cloquence$ files.
+Therefore, by re-using an existing, widely-used grammar, we are able to satisfy many of the design criteria described in Section \ref{syntax} and the adoption criteria described in Section \ref{tools} without significant development effort.
+
+\subsection{Semantics}
+Unlike Python, $\cloquence$ is a compiled, statically-typed language.
+The extension used in Figure \ref{map} implements the full OpenCL type system, which includes features inherited from C99 like pointer arithmetic and OpenCL-specific constructs like vector types.
+In fact, compilation of Figure \ref{map} will produce code semantically equivalent to, and with exactly the same performance profile as, manually-written OpenCL code.
+Compilation produces OpenCL source code, so $\cloquence$, when used with this extension, also inherits OpenCL's portability profile as well, satisfying criteria described in Section \ref{portability}.
+
+In OpenCL itself, however, there is no way to write a function like \verb|map|.
+As described in Section \ref{hof}, OpenCL lacks support for higher-order functions or polymorphism (over the types of \verb|in| and \verb|out| in this case.) To map a function over an array, developers must create a separate function for every combination of argument types and for every transfer function.
+Figure 2 shows an OpenCL kernel that implements a map specifically for the \verb|sin| function applied to arrays of doubles in global memory.
+Note that because OpenCL does not support templates or function pointers, this is the most general solution possible without explicit code generation.
+However, even in CUDA, which does support templates, there is greater syntactic overhead for specializing a function with particular types than in $\cloquence$, where it happens implicitly.
+
 \begin{figure}
@@
 
-\subsection{Semantics}
-$\cloquence$ is a compiled, statically-typed language.
-The extension used in Figure \ref{map} implements the full OpenCL type system, which includes features inherited from C99 like pointer arithmetic and OpenCL-specific constructs like vector types.
-In fact, compilation of Figure \ref{map} will produce code semantically equivalent, and with the same performance profile, as manually-written OpenCL code.
-Compilation produces OpenCL source code, so $\cloquence$, when used with this extension, inherits OpenCL's portability profile as well, as described in Section \ref{portability}.
-
-In OpenCL itself, however, there is no way to write a function like \verb|map|.
-As described in Section \ref{hof}, OpenCL lacks support for higher-order functions or polymorphism (over the types of \verb|in| and \verb|out| in this case.) To map a function over an array, developers must create a separate function for every combination of argument types and for every transfer function.
-Figure 2 shows an OpenCL kernel that implements a map specifically for the \verb|sin| function applied to arrays of doubles in global memory.
-Note that because OpenCL does not support templates or function pointers, this is the most general solution possible without explicit code generation.
-However, even in CUDA, which does support templates, there is greater syntactic overhead for specializing a function with particular types than in $\cloquence$, where it happens implicitly.
-
-To support this much more concise form for specifying kernels, $\cloquence$ uses techniques pioneered by functional programming languages. We briefly describe these below.
+To support this much more concise form for specifying kernels, $\cloquence$ uses variants of techniques originally  pioneered by functional programming languages. We briefly describe these below.
 
@@
 It can, however, result in ``code bloat'', since a new copy of the higher-order function is created for each unique function argument passed to it.
-First-class functions can theoretically be simulated using an auxiliary lookup function in OpenCL, and we leave clean support for this in $\cloquence$ as open future work.
+First-class functions can theoretically be simulated using an auxiliary lookup function in OpenCL, and we leave clean support for this in $\cloquence$ as straightforward future work.
 
@@
 Each neuron is modeled using one or more ordinary differential equations.
-These equations capture the dynamics of physically important quantities like the cell's membrane potential or the conductance across various kinds of ion channels and can take many forms \cite{neurobook}.
+These equations capture the dynamics of physically important quantities like the cell's membrane potential or the conductance across various kinds of ion channels and can take many forms.
 Single simulations can contain from hundreds to tens of millions of neurons each, depending on the specific problem being studied.
@@
 
-The research group we discuss here (of which the first author was a member) was studying a problem that required running up to 1,000 realizations of a network of between 4,000 and 10,000 neurons each.
+The research group we discuss here (of which the first author was formerly a member) was studying a problem that required running up to 1,000 realizations of a network of between 4,000 and 10,000 neurons each.
 An initial solution to this problem used the Brian framework, written in Python, to conduct these simulations on a CPU.
-Brian was selected because it allowed the structure of the  simulation to be specified in modular and straightforward manner.
+Brian was selected because it allowed the structure of the  simulation to be specified in a modular and straightforward manner.
 This solution required between 60 and 70 minutes to conduct the simulations and up to 8 hours to analyze the data each time a parameter of the simulation was modified.
@@
 If interleaved analysis needed to be conducted as well, it would be specified in the same way.
-
-Implementations of these classes do not evaluate the simulation logic directly, but rather contain methods that generate $\cloquence$ source code for insertion at various points, called {\it hooks}, in the final simulation kernel.
-The hook that code is inserted into is determined by the method name, and code can be inserted into any hook defined anywhere upstream in the simulation tree.
-New hooks can also be defined in these methods and these become available for use by child nodes.
-Figure \ref{impl} shows an example of a class that inserts code in the \verb|model_code| hook and defines several new hooks.
-This protocol is closely related to the notion of {\it frame-oriented programming}.
-Although highly modular, this strategy avoids the performance penalties associated with standard object-oriented methodologies via code generation.
-
-Compared to a similar protocol targeting OpenCL directly, the required code generation logic is significantly simpler because it enables classes like \verb|StateVariable| to be written generically for all types of state variables, without carrying extra parameters and {\it ad hoc} logic to extract and compute the result types of generated expressions.
-Moreover, because types are first-class objects in the metalanguage, they can be examined during the memory allocation step to enable features like fully-automatic parallelization of multiple realizations across one or more devices, a major feature of {\sf cl.egans} that competing frameworks cannot easily offer.
-
-Once the kernel has been generated and memory has been allocated, the simulation can be executed directly from Python using the bindings described in Section \ref{direct}.
-The results of this simulation are immediately available to the Python code following the simulation and can be visualized and further analyzed using standard tools.
-Once the computations are complete, the Python garbage collector is able to handle deallocation of GPU memory automatically (a feature of the underlying \verb|pyopencl| library \cite{klockner2011pycuda}.)
-
-Using this $\cloquence$-based framework, the benefits of the Brian-based workflow were recovered without the  corresponding decrease in performance relative to the previous CUDA-based solution, leading ultimately to a satisfying solution for the group conducting this research.
-
 \begin{figure}
@@
 \end{figure}
+
+Implementations of these classes do not evaluate the simulation logic directly, but rather contain methods that generate $\cloquence$ source code for insertion at various points, called {\it hooks}, in the final simulation kernel.
+The hook that code is inserted into is determined by the method name, and code can be inserted into any hook defined by any node upstream in the simulation tree.
+New hooks can also be defined in these methods and these become available for use by child nodes.
+Figure \ref{impl} shows an example of a class that inserts code in the \verb|model_code| hook and defines several new hooks.
+This protocol is closely related to the notion of {\it frame-oriented programming}.
+Although highly modular, this strategy avoids the performance penalties associated with standard object-oriented methodologies via code generation.
 
@@
 
+\subsection{Benefits}
+Compared to a similar protocol targeting OpenCL directly, the required code generation logic is significantly simpler because it enables classes like \verb|StateVariable| to be written generically for all types of state variables, without carrying extra parameters and {\it ad hoc} logic to extract and compute the result types of generated expressions.
+Moreover, because types are first-class objects in the metalanguage, they can be examined during the memory allocation step to enable features like fully-automatic parallelization of multiple realizations across one or more devices, a major feature of {\sf cl.egans} that competing frameworks cannot easily offer.
+
+Once the kernel has been generated and memory has been allocated, the simulation can be executed directly from Python using the bindings described in Section \ref{direct}.
+The results of this simulation are immediately available to the Python code following the simulation and can be visualized and further analyzed using standard tools.
+Once the computations are complete, the Python garbage collector is able to handle deallocation of GPU memory automatically (a feature of the underlying \verb|pyopencl| library \cite{klockner2011pycuda}.)
+
+Using this $\cloquence$-based framework, the benefits of the Brian-based workflow were recovered without the  corresponding decrease in performance relative to the previous CUDA-based solution, leading ultimately to a satisfying solution for the group conducting this research.
+
 \section{Conclusion}
@@
 
-% An example of a floating figure using the graphicx package.
-% Note that \label must occur AFTER (or within) \caption.
-% For figures, \caption should occur after the \includegraphics.
-% Note that IEEEtran v1.7 and later has special internal code that
-% is designed to preserve the operation of \label within \caption
-% even when the captionsoff option is in effect. However, because
-% of issues like this, it may be the safest practice to put all your
-% \label just after \caption rather than within \caption{}.
 %
-% Reminder: the "draftcls" or "draftclsnofoot", not "draft", class
-% option should be used if it is desired that the figures are to be
-% displayed while in draft mode.
+
 %
-%\begin{figure}[!t]
-%\centering
-%\includegraphics[width=2.5in]{myfigure}
-% where an .eps filename suffix will be assumed under latex, 
-% and a .pdf suffix will be assumed for pdflatex; or what has been declared
-% via \DeclareGraphicsExtensions.
-%\caption{Simulation Results}
-%\label{fig_sim}
-%\end{figure}
-
-% Note that IEEE typically puts floats only at the top, even when this
-% results in a large percentage of a column being occupied by floats.
-
-% An example of a double column floating figure using two subfigures.
-% (The subfig.sty package must be loaded for this to work.)
-% The subfigure \label commands are set within each subfloat command, the
-% \label for the overall figure must come after \caption.
-% \hfil must be used as a separator to get equal spacing.
-% The subfigure.sty package works much the same way, except \subfigure is
-% used instead of \subfloat.
+% ---- Bibliography ----
+\bibliographystyle{IEEEtran}
+\bibliography{../research}
 %
-%\begin{figure*}[!t]
-%\centerline{\subfloat[Case I]\includegraphics[width=2.5in]{subfigcase1}%
-%\label{fig_first_case}}
-%\hfil
-%\subfloat[Case II]{\includegraphics[width=2.5in]{subfigcase2}%
-%\label{fig_second_case}}}
-%\caption{Simulation results}
-%\label{fig_sim}
-%\end{figure*}
+%\begin{thebibliography}{}
 %
-% Note that often IEEE papers with subfigures do not employ subfigure
-% captions (using the optional argument to \subfloat), but instead will
-% reference/describe all of them (a), (b), etc., within the main caption.
-
-% An example of a floating table. Note that, for IEEE style tables, the 
-% \caption command should come BEFORE the table. Table text will default to
-% \footnotesize as IEEE normally uses this smaller font for tables.
-% The \label must come after \caption as always.
+%\bibitem[1980]{2clar:eke}
+%Clarke, F., Ekeland, I.:
+%Nonlinear oscillations and
+%boundary-value problems for Hamiltonian systems.
+%Arch. Rat. Mech. Anal. {\bf 78} (1982) 315--333
 %
-%\begin{table}[!t]
-%% increase table row spacing, adjust to taste
-%\renewcommand{\arraystretch}{1.3}
-% if using array.sty, it might be a good idea to tweak the value of
-% \extrarowheight as needed to properly center the text within the cells
-%\caption{An Example of a Table}
-%\label{table_example}
-%\centering
-%% Some packages, such as MDW tools, offer better commands for making tables
-%% than the plain LaTeX2e tabular which is used here.
-%\begin{tabular}{|c||c|}
-%\hline
-%One & Two\\
-%\hline
-%Three & Four\\
-%\hline
-%\end{tabular}
-%\end{table}
-
-% Note that IEEE does not put floats in the very first column - or typically
-% anywhere on the first page for that matter. Also, in-text middle ("here")
-% positioning is not used. Most IEEE journals/conferences use top floats
-% exclusively. Note that, LaTeX2e, unlike IEEE journals/conferences, places
-% footnotes above bottom floats. This can be corrected via the \fnbelowfloat
-% command of the stfloats package.
-
-% conference papers do not normally have an appendix
-
-% use section* for acknowledgement
-% trigger a \newpage just before the given reference
-% number - used to balance the columns on the last page
-% adjust value as needed - may need to be readjusted if
-% the document is modified later
-%\IEEEtriggeratref{8}
-% The "triggered" command can be changed if desired:
-%\IEEEtriggercmd{\enlargethispage{-5in}}
-
-% references section
-
-% can use a bibliography generated by BibTeX as a .bbl file
-% BibTeX documentation can be easily obtained at:
-% http://www.ctan.org/tex-archive/biblio/bibtex/contrib/doc/
-% The IEEEtran BibTeX style support page is at:
-% http://www.michaelshell.org/tex/ieeetran/bibtex/
-%\bibliographystyle{IEEEtran}
-% argument is your BibTeX string definitions and bibliography database(s)
-%\bibliography{IEEEabrv,../bib/paper}
+%\bibitem[1981]{2clar:eke:2}
+%Clarke, F., Ekeland, I.:
+%Solutions p\'{e}riodiques, du
+%p\'{e}riode donn\'{e}e, des \'{e}quations hamiltoniennes.
+%Note CRAS Paris {\bf 287} (1978) 1013--1015
 %
-% <OR> manually copy in the resultant .bbl file
-% set second argument of \begin to the number of references
-% (used to reserve space for the reference number labels box)
-\bibliographystyle{IEEEtran}
-\bibliography{research}
+%\bibitem[1982]{2mich:tar}
+%Michalek, R., Tarantello, G.:
+%Subharmonic solutions with prescribed minimal
+%period for nonautonomous Hamiltonian systems.
+%J. Diff. Eq. {\bf 72} (1988) 28--55
 %
-%\begin{thebibliography}{1}
+%\bibitem[1983]{2tar}
+%Tarantello, G.:
+%Subharmonic solutions for Hamiltonian
+%systems via a $\bbbz_{p}$ pseudoindex theory.
+%Annali di Matematica Pura (to appear)
 %
-%\bibitem{IEEEhowto:kopka}
-%H.~Kopka and P.~W. Daly, \emph{A Guide to \LaTeX}, 3rd~ed.\hskip 1em plus
-%  0.5em minus 0.4em\relax Harlow, England: Addison-Wesley, 1999.
+%\bibitem[1985]{2rab}
+%Rabinowitz, P.:
+%On subharmonic solutions of a Hamiltonian system.
+%Comm. Pure Appl. Math. {\bf 33} (1980) 609--633
 %
 %\end{thebibliography}
-
-% that's all folks
+%\clearpage
+%\addtocmark[2]{Author Index} % additional numbered TOC entry
+%\renewcommand{\indexname}{Author Index}
+%\printindex
+%\clearpage
+%\addtocmark[2]{Subject Index} % additional numbered TOC entry
+%\markboth{Subject Index}{Subject Index}
+%\renewcommand{\indexname}{Subject Index}
+%\input{subjidx.ind}
 \end{document}

### 2014-11-25 ecoop12/llncs2e/llncs.tex
Updated proposal

% This is LLNCS.DEM the demonstration file of
% the LaTeX macro package from Springer-Verlag
% for Lecture Notes in Computer Science,
% version 2.2 for LaTeX2e
%
\documentclass{llncs}
%
\usepackage{makeidx}  % allows for indexgeneration
%
\begin{document}
%
\frontmatter          % for the preliminaries
%
\pagestyle{headings}  % switches on printing of running heads
\addtocmark{Hamiltonian Mechanics} % additional mark in the TOC
%
\mainmatter              % start of the contributions
%
\title{Hamiltonian Mechanics unter besonderer Ber\"ucksichtigung der
h\"ohreren Lehranstalten}
%
\titlerunning{Hamiltonian Mechanics}  % abbreviated title (for running head)
%                                     also used for the TOC unless
%                                     \toctitle is used
%
\author{Ivar Ekeland\inst{1} \and Roger Temam\inst{2}
Jeffrey Dean \and David Grove \and Craig Chambers \and Kim~B.~Bruce \and
Elsa Bertino}
%
\authorrunning{Ivar Ekeland et al.}   % abbreviated author list (for running head)
%
%%%% modified list of authors for the TOC (add the affiliations)
\tocauthor{Ivar Ekeland (Princeton University),
Roger Temam (Universit\'{e} de Paris-Sud),
Jeffrey Dean, David Grove, Craig Chambers (Universit\`a di Geova),
Kim B. Bruce (Stanford University),
Elisa Bertino (Digita Research Center)}
%
\institute{Princeton University, Princeton NJ 08544, USA,\\
\email{I.Ekeland@princeton.edu},\\ WWW home page:
\texttt{http://users/\homedir iekeland/web/welcome.html}
\and
Universit\'{e} de Paris-Sud,
Laboratoire d'Analyse Num\'{e}rique, B\^{a}timent 425,\\
F-91405 Orsay Cedex, France}

\maketitle              % typeset the title of the contribution

\begin{abstract}
The abstract should summarize the contents of the paper
using at least 70 and at most 150 words. It will be set in 9-point
font size and be inset 1.0 cm from the right and left margins.
There will be two blank lines before and after the Abstract. \dots
\end{abstract}
%
\section{Fixed-Period Problems: The Sublinear Case}
%
With this chapter, the preliminaries are over, and we begin the search
for periodic solutions to Hamiltonian systems. All this will be done in
the convex case; that is, we shall study the boundary-value problem
\begin{eqnarray*}
  \dot{x}&=&JH' (t,x)\\
  x(0) &=& x(T)
\end{eqnarray*}
with $H(t,\cdot)$ a convex function of $x$, going to $+\infty$ when
$\left\|x\right\| \to \infty$.

%
\subsection{Autonomous Systems}
%
In this section, we will consider the case when the Hamiltonian $H(x)$
is autonomous. For the sake of simplicity, we shall also assume that it
is $C^{1}$.

We shall first consider the question of nontriviality, within the
general framework of
$\left(A_{\infty},B_{\infty}\right)$-subquadratic Hamiltonians. In
the second subsection, we shall look into the special case when $H$ is
$\left(0,b_{\infty}\right)$-subquadratic,
and we shall try to derive additional information.
%
\subsubsection{The General Case: Nontriviality.}
%
We assume that $H$ is
$\left(A_{\infty},B_{\infty}\right)$-sub\-qua\-dra\-tic at infinity,
for some constant symmetric matrices $A_{\infty}$ and $B_{\infty}$,
with $B_{\infty}-A_{\infty}$ positive definite. Set:
\begin{eqnarray}
\gamma :&=&{\rm smallest\ eigenvalue\ of}\ \ B_{\infty} - A_{\infty} \\
  \lambda : &=& {\rm largest\ negative\ eigenvalue\ of}\ \
  J \frac{d}{dt} +A_{\infty}\ .
\end{eqnarray}

Theorem~\ref{ghou:pre} tells us that if $\lambda +\gamma < 0$, the
boundary-value problem:
\begin{equation}
\begin{array}{rcl}
  \dot{x}&=&JH' (x)\\
  x(0)&=&x (T)
\end{array}
\end{equation}
has at least one solution
$\overline{x}$, which is found by minimizing the dual
action functional:
\begin{equation}
  \psi (u) = \int_{o}^{T} \left[\frac{1}{2}
  \left(\Lambda_{o}^{-1} u,u\right) + N^{\ast} (-u)\right] dt
\end{equation}
on the range of $\Lambda$, which is a subspace $R (\Lambda)_{L}^{2}$
with finite codimension. Here
\begin{equation}
  N(x) := H(x) - \frac{1}{2} \left(A_{\infty} x,x\right)
\end{equation}
is a convex function, and
\begin{equation}
  N(x) \le \frac{1}{2}
  \left(\left(B_{\infty} - A_{\infty}\right) x,x\right)
  + c\ \ \ \forall x\ .
\end{equation}

%
\begin{proposition}
Assume $H'(0)=0$ and $ H(0)=0$. Set:
\begin{equation}
  \delta := \liminf_{x\to 0} 2 N (x) \left\|x\right\|^{-2}\ .
  \label{eq:one}
\end{equation}

If $\gamma < - \lambda < \delta$,
the solution $\overline{u}$ is non-zero:
\begin{equation}
  \overline{x} (t) \ne 0\ \ \ \forall t\ .
\end{equation}
\end{proposition}
%
\begin{proof}
Condition (\ref{eq:one}) means that, for every
$\delta ' > \delta$, there is some $\varepsilon > 0$ such that
\begin{equation}
  \left\|x\right\| \le \varepsilon \Rightarrow N (x) \le
  \frac{\delta '}{2} \left\|x\right\|^{2}\ .
\end{equation}

It is an exercise in convex analysis, into which we shall not go, to
show that this implies that there is an $\eta > 0$ such that
\begin{equation}
  f\left\|x\right\| \le \eta
  \Rightarrow N^{\ast} (y) \le \frac{1}{2\delta '}
  \left\|y\right\|^{2}\ .
  \label{eq:two}
\end{equation}

\begin{figure}
\vspace{2.5cm}
\caption{This is the caption of the figure displaying a white eagle and
a white horse on a snow field}
\end{figure}

Since $u_{1}$ is a smooth function, we will have
$\left\|hu_{1}\right\|_\infty \le \eta$
for $h$ small enough, and inequality (\ref{eq:two}) will hold,
yielding thereby:
\begin{equation}
  \psi (hu_{1}) \le \frac{h^{2}}{2}
  \frac{1}{\lambda} \left\|u_{1} \right\|_{2}^{2} + \frac{h^{2}}{2}
  \frac{1}{\delta '} \left\|u_{1}\right\|^{2}\ .
\end{equation}

If we choose $\delta '$ close enough to $\delta$, the quantity
$\left(\frac{1}{\lambda} + \frac{1}{\delta '}\right)$
will be negative, and we end up with
\begin{equation}
  \psi (hu_{1}) < 0\ \ \ \ \ {\rm for}\ \ h\ne 0\ \ {\rm small}\ .
\end{equation}

On the other hand, we check directly that $\psi (0) = 0$. This shows
that 0 cannot be a minimizer of $\psi$, not even a local one.
So $\overline{u} \ne 0$ and
$\overline{u} \ne \Lambda_{o}^{-1} (0) = 0$. \qed
\end{proof}
%
\begin{corollary}
Assume $H$ is $C^{2}$ and
$\left(a_{\infty},b_{\infty}\right)$-subquadratic at infinity. Let
$\xi_{1},\allowbreak\dots,\allowbreak\xi_{N}$  be the
equilibria, that is, the solutions of $H' (\xi ) = 0$.
Denote by $\omega_{k}$
the smallest eigenvalue of $H'' \left(\xi_{k}\right)$, and set:
\begin{equation}
  \omega : = {\rm Min\,} \left\{\omega_{1},\dots,\omega_{k}\right\}\ .
\end{equation}
If:
\begin{equation}
  \frac{T}{2\pi} b_{\infty} <
  - E \left[- \frac{T}{2\pi}a_{\infty}\right] <
  \frac{T}{2\pi}\omega
  \label{eq:three}
\end{equation}
then minimization of $\psi$ yields a non-constant $T$-periodic solution
$\overline{x}$.
\end{corollary}
%

We recall once more that by the integer part $E [\alpha ]$ of
$\alpha \in \bbbr$, we mean the $a\in \bbbz$
such that $a< \alpha \le a+1$. For instance,
if we take $a_{\infty} = 0$, Corollary 2 tells
us that $\overline{x}$ exists and is
non-constant provided that:

\begin{equation}
  \frac{T}{2\pi} b_{\infty} < 1 < \frac{T}{2\pi}
\end{equation}
or
\begin{equation}
  T\in \left(\frac{2\pi}{\omega},\frac{2\pi}{b_{\infty}}\right)\ .
  \label{eq:four}
\end{equation}

%
\begin{proof}
The spectrum of $\Lambda$ is $\frac{2\pi}{T} \bbbz +a_{\infty}$. The
largest negative eigenvalue $\lambda$ is given by
$\frac{2\pi}{T}k_{o} +a_{\infty}$,
where
\begin{equation}
  \frac{2\pi}{T}k_{o} + a_{\infty} < 0
  \le \frac{2\pi}{T} (k_{o} +1) + a_{\infty}\ .
\end{equation}
Hence:
\begin{equation}
  k_{o} = E \left[- \frac{T}{2\pi} a_{\infty}\right] \ .
\end{equation}

The condition $\gamma < -\lambda < \delta$ now becomes:
\begin{equation}
  b_{\infty} - a_{\infty} <
  - \frac{2\pi}{T} k_{o} -a_{\infty} < \omega -a_{\infty}
\end{equation}
which is precisely condition (\ref{eq:three}).\qed
\end{proof}
%

\begin{lemma}
Assume that $H$ is $C^{2}$ on $\bbbr^{2n} \setminus \{ 0\}$ and
that $H'' (x)$ is non-de\-gen\-er\-ate for any $x\ne 0$. Then any local
minimizer $\widetilde{x}$ of $\psi$ has minimal period $T$.
\end{lemma}
%
\begin{proof}
We know that $\widetilde{x}$, or
$\widetilde{x} + \xi$ for some constant $\xi
\in \bbbr^{2n}$, is a $T$-periodic solution of the Hamiltonian system:
\begin{equation}
  \dot{x} = JH' (x)\ .
\end{equation}

There is no loss of generality in taking $\xi = 0$. So
$\psi (x) \ge \psi (\widetilde{x} )$
for all $\widetilde{x}$ in some neighbourhood of $x$ in
$W^{1,2} \left(\bbbr / T\bbbz ; \bbbr^{2n}\right)$.

But this index is precisely the index
$i_{T} (\widetilde{x} )$ of the $T$-periodic
solution $\widetilde{x}$ over the interval
$(0,T)$, as defined in Sect.~2.6. So
\begin{equation}
  i_{T} (\widetilde{x} ) = 0\ .
  \label{eq:five}
\end{equation}

Now if $\widetilde{x}$ has a lower period, $T/k$ say,
we would have, by Corollary 31:
\begin{equation}
  i_{T} (\widetilde{x} ) =
  i_{kT/k}(\widetilde{x} ) \ge
  ki_{T/k} (\widetilde{x} ) + k-1 \ge k-1 \ge 1\ .
\end{equation}

This would contradict (\ref{eq:five}), and thus cannot happen.\qed
\end{proof}
%
\paragraph{Notes and Comments.}
The results in this section are a
refined version of \cite{clar:eke};
the minimality result of Proposition
14 was the first of its kind.

To understand the nontriviality conditions, such as the one in formula
(\ref{eq:four}), one may think of a one-parameter family
$x_{T}$, $T\in \left(2\pi\omega^{-1}, 2\pi b_{\infty}^{-1}\right)$
of periodic solutions, $x_{T} (0) = x_{T} (T)$,
with $x_{T}$ going away to infinity when $T\to 2\pi \omega^{-1}$,
which is the period of the linearized system at 0.

\begin{table}
\caption{This is the example table taken out of {\it The
\TeX{}book,} p.\,246}
\begin{center}
\begin{tabular}{r@{\quad}rl}
\hline
\multicolumn{1}{l}{\rule{0pt}{12pt}
                   Year}&\multicolumn{2}{l}{World population}\\[2pt]
\hline\rule{0pt}{12pt}
8000 B.C.  &     5,000,000& \\
  50 A.D.  &   200,000,000& \\
1650 A.D.  &   500,000,000& \\
1945 A.D.  & 2,300,000,000& \\
1980 A.D.  & 4,400,000,000& \\[2pt]
\hline
\end{tabular}
\end{center}
\end{table}
%
\begin{theorem} [Ghoussoub-Preiss]\label{ghou:pre}
Assume $H(t,x)$ is
$(0,\varepsilon )$-subquadratic at
infinity for all $\varepsilon > 0$, and $T$-periodic in $t$
\begin{equation}
  H (t,\cdot )\ \ \ \ \ {\rm is\ convex}\ \ \forall t
\end{equation}
\begin{equation}
  H (\cdot ,x)\ \ \ \ \ {\rm is}\ \ T{\rm -periodic}\ \ \forall x
\end{equation}
\begin{equation}
  H (t,x)\ge n\left(\left\|x\right\|\right)\ \ \ \ \
  {\rm with}\ \ n (s)s^{-1}\to \infty\ \ {\rm as}\ \ s\to \infty
\end{equation}
\begin{equation}
  \forall \varepsilon > 0\ ,\ \ \ \exists c\ :\
  H(t,x) \le \frac{\varepsilon}{2}\left\|x\right\|^{2} + c\ .
\end{equation}

Assume also that $H$ is $C^{2}$, and $H'' (t,x)$ is positive definite
everywhere. Then there is a sequence $x_{k}$, $k\in \bbbn$, of
$kT$-periodic solutions of the system
\begin{equation}
  \dot{x} = JH' (t,x)
\end{equation}
such that, for every $k\in \bbbn$, there is some $p_{o}\in\bbbn$ with:
\begin{equation}
  p\ge p_{o}\Rightarrow x_{pk} \ne x_{k}\ .
\end{equation}
\qed
\end{theorem}
%
\begin{example} [{{\rm External forcing}}]
Consider the system:
\begin{equation}
  \dot{x} = JH' (x) + f(t)
\end{equation}
where the Hamiltonian $H$ is
$\left(0,b_{\infty}\right)$-subquadratic, and the
forcing term is a distribution on the circle:
\begin{equation}
  f = \frac{d}{dt} F + f_{o}\ \ \ \ \
  {\rm with}\ \ F\in L^{2} \left(\bbbr / T\bbbz; \bbbr^{2n}\right)\ ,
\end{equation}
where $f_{o} : = T^{-1}\int_{o}^{T} f (t) dt$. For instance,
\begin{equation}
  f (t) = \sum_{k\in \bbbn} \delta_{k} \xi\ ,
\end{equation}
where $\delta_{k}$ is the Dirac mass at $t= k$ and
$\xi \in \bbbr^{2n}$ is a
constant, fits the prescription. This means that the system
$\dot{x} = JH' (x)$ is being excited by a
series of identical shocks at interval $T$.
\end{example}
%
\begin{definition}
Let $A_{\infty} (t)$ and $B_{\infty} (t)$ be symmetric
operators in $\bbbr^{2n}$, depending continuously on
$t\in [0,T]$, such that
$A_{\infty} (t) \le B_{\infty} (t)$ for all $t$.

A Borelian function
$H: [0,T]\times \bbbr^{2n} \to \bbbr$
is called
$\left(A_{\infty} ,B_{\infty}\right)$-{\it subquadratic at infinity}
if there exists a function $N(t,x)$ such that:
\begin{equation}
  H (t,x) = \frac{1}{2} \left(A_{\infty} (t) x,x\right) + N(t,x)
\end{equation}
\begin{equation}
  \forall t\ ,\ \ \ N(t,x)\ \ \ \ \
  {\rm is\ convex\ with\  respect\  to}\ \ x
\end{equation}
\begin{equation}
  N(t,x) \ge n\left(\left\|x\right\|\right)\ \ \ \ \
  {\rm with}\ \ n(s)s^{-1}\to +\infty\ \ {\rm as}\ \ s\to +\infty
\end{equation}
\begin{equation}
  \exists c\in \bbbr\ :\ \ \ H (t,x) \le
  \frac{1}{2} \left(B_{\infty} (t) x,x\right) + c\ \ \ \forall x\ .
\end{equation}

If $A_{\infty} (t) = a_{\infty} I$ and
$B_{\infty} (t) = b_{\infty} I$, with
$a_{\infty} \le b_{\infty} \in \bbbr$,
we shall say that $H$ is
$\left(a_{\infty},b_{\infty}\right)$-subquadratic
at infinity. As an example, the function
$\left\|x\right\|^{\alpha}$, with
$1\le \alpha < 2$, is $(0,\varepsilon )$-subquadratic at infinity
for every $\varepsilon > 0$. Similarly, the Hamiltonian
\begin{equation}
H (t,x) = \frac{1}{2} k \left\|k\right\|^{2} +\left\|x\right\|^{\alpha}
\end{equation}
is $(k,k+\varepsilon )$-subquadratic for every $\varepsilon > 0$.
Note that, if $k<0$, it is not convex.
\end{definition}
%

\paragraph{Notes and Comments.}
The first results on subharmonics were
obtained by Rabinowitz in \cite{rab}, who showed the existence of
infinitely many subharmonics both in the subquadratic and superquadratic
case, with suitable growth conditions on $H'$. Again the duality
approach enabled Clarke and Ekeland in \cite{clar:eke:2} to treat the
same problem in the convex-subquadratic case, with growth conditions on
$H$ only.

Recently, Michalek and Tarantello (see \cite{mich:tar} and \cite{tar})
have obtained lower bound on the number of subharmonics of period $kT$,
based on symmetry considerations and on pinching estimates, as in
Sect.~5.2 of this article.

%
% ---- Bibliography ----
%
\begin{thebibliography}{5}
%
\bibitem {clar:eke}
Clarke, F., Ekeland, I.:
Nonlinear oscillations and
boundary-value problems for Hamiltonian systems.
Arch. Rat. Mech. Anal. {\bf 78} (1982) 315--333

\bibitem {clar:eke:2}
Clarke, F., Ekeland, I.:
Solutions p\'{e}riodiques, du
p\'{e}riode donn\'{e}e, des \'{e}quations hamiltoniennes.
Note CRAS Paris {\bf 287} (1978) 1013--1015

\bibitem {mich:tar}
Michalek, R., Tarantello, G.:
Subharmonic solutions with prescribed minimal
period for nonautonomous Hamiltonian systems.
J. Diff. Eq. {\bf 72} (1988) 28--55

\bibitem {tar}
Tarantello, G.:
Subharmonic solutions for Hamiltonian
systems via a $\bbbz_{p}$ pseudoindex theory.
Annali di Matematica Pura (to appear)

\bibitem {rab}
Rabinowitz, P.:
On subharmonic solutions of a Hamiltonian system.
Comm. Pure Appl. Math. {\bf 33} (1980) 609--633

\end{thebibliography}

%
% second contribution with nearly identical text,
% slightly changed contribution head (all entries
% appear as defaults), and modified bibliography
%
\title{Hamiltonian Mechanics2}

\author{Ivar Ekeland\inst{1} \and Roger Temam\inst{2}}

\institute{Princeton University, Princeton NJ 08544, USA
\and
Universit\'{e} de Paris-Sud,
Laboratoire d'Analyse Num\'{e}rique, B\^{a}timent 425,\\
F-91405 Orsay Cedex, France}

\maketitle
%
% Modify the bibliography environment to call for the author-year
% system. This is done normally with the citeauthoryear option
% for a particular contribution.
\makeatletter
\renewenvironment{thebibliography}[1]
     {\section*{\refname}
      \small
      \list{}%
           {\settowidth\labelwidth{}%
            \leftmargin\parindent
            \itemindent=-\parindent
            \labelsep=\z@
            \if@openbib
              \advance\leftmargin\bibindent
              \itemindent -\bibindent
              \listparindent \itemindent
              \parsep \z@
            \fi
            \usecounter{enumiv}%
            \let\p@enumiv\@empty
            \renewcommand\theenumiv{}}%
      \if@openbib
        \renewcommand\newblock{\par}%
      \else
        \renewcommand\newblock{\hskip .11em \@plus.33em \@minus.07em}%
      \fi
      \sloppy\clubpenalty4000\widowpenalty4000%
      \sfcode`\.=\@m}
     {\def\@noitemerr
       {\@latex@warning{Empty `thebibliography' environment}}%
      \endlist}
      \def\@cite#1{#1}%
      \def\@lbibitem[#1]#2{\item[]\if@filesw
        {\def\protect##1{\string ##1\space}\immediate
      \write\@auxout{\string\bibcite{#2}{#1}}}\fi\ignorespaces}
\makeatother
%
\begin{abstract}
The abstract should summarize the contents of the paper
using at least 70 and at most 150 words. It will be set in 9-point
font size and be inset 1.0 cm from the right and left margins.
There will be two blank lines before and after the Abstract. \dots
\end{abstract}
%
\section{Fixed-Period Problems: The Sublinear Case}
%
With this chapter, the preliminaries are over, and we begin the search
for periodic solutions to Hamiltonian systems. All this will be done in
the convex case; that is, we shall study the boundary-value problem
\begin{eqnarray*}
  \dot{x}&=&JH' (t,x)\\
  x(0) &=& x(T)
\end{eqnarray*}
with $H(t,\cdot)$ a convex function of $x$, going to $+\infty$ when
$\left\|x\right\| \to \infty$.

%
\subsection{Autonomous Systems}
%
In this section, we will consider the case when the Hamiltonian $H(x)$
is autonomous. For the sake of simplicity, we shall also assume that it
is $C^{1}$.

We shall first consider the question of nontriviality, within the
general framework of
$\left(A_{\infty},B_{\infty}\right)$-subquadratic Hamiltonians. In
the second subsection, we shall look into the special case when $H$ is
$\left(0,b_{\infty}\right)$-subquadratic,
and we shall try to derive additional information.
%
\subsubsection{The General Case: Nontriviality.}
%
We assume that $H$ is
$\left(A_{\infty},B_{\infty}\right)$-sub\-qua\-dra\-tic at infinity,
for some constant symmetric matrices $A_{\infty}$ and $B_{\infty}$,
with $B_{\infty}-A_{\infty}$ positive definite. Set:
\begin{eqnarray}
\gamma :&=&{\rm smallest\ eigenvalue\ of}\ \ B_{\infty} - A_{\infty} \\
  \lambda : &=& {\rm largest\ negative\ eigenvalue\ of}\ \
  J \frac{d}{dt} +A_{\infty}\ .
\end{eqnarray}

Theorem 21 tells us that if $\lambda +\gamma < 0$, the boundary-value
problem:
\begin{equation}
\begin{array}{rcl}
  \dot{x}&=&JH' (x)\\
  x(0)&=&x (T)
\end{array}
\end{equation}
has at least one solution
$\overline{x}$, which is found by minimizing the dual
action functional:
\begin{equation}
  \psi (u) = \int_{o}^{T} \left[\frac{1}{2}
  \left(\Lambda_{o}^{-1} u,u\right) + N^{\ast} (-u)\right] dt
\end{equation}
on the range of $\Lambda$, which is a subspace $R (\Lambda)_{L}^{2}$
with finite codimension. Here
\begin{equation}
  N(x) := H(x) - \frac{1}{2} \left(A_{\infty} x,x\right)
\end{equation}
is a convex function, and
\begin{equation}
  N(x) \le \frac{1}{2}
  \left(\left(B_{\infty} - A_{\infty}\right) x,x\right)
  + c\ \ \ \forall x\ .
\end{equation}

%
\begin{proposition}
Assume $H'(0)=0$ and $ H(0)=0$. Set:
\begin{equation}
  \delta := \liminf_{x\to 0} 2 N (x) \left\|x\right\|^{-2}\ .
  \label{2eq:one}
\end{equation}

If $\gamma < - \lambda < \delta$,
the solution $\overline{u}$ is non-zero:
\begin{equation}
  \overline{x} (t) \ne 0\ \ \ \forall t\ .
\end{equation}
\end{proposition}
%
\begin{proof}
Condition (\ref{2eq:one}) means that, for every
$\delta ' > \delta$, there is some $\varepsilon > 0$ such that
\begin{equation}
  \left\|x\right\| \le \varepsilon \Rightarrow N (x) \le
  \frac{\delta '}{2} \left\|x\right\|^{2}\ .
\end{equation}

It is an exercise in convex analysis, into which we shall not go, to
show that this implies that there is an $\eta > 0$ such that
\begin{equation}
  f\left\|x\right\| \le \eta
  \Rightarrow N^{\ast} (y) \le \frac{1}{2\delta '}
  \left\|y\right\|^{2}\ .
  \label{2eq:two}
\end{equation}

\begin{figure}
\vspace{2.5cm}
\caption{This is the caption of the figure displaying a white eagle and
a white horse on a snow field}
\end{figure}

Since $u_{1}$ is a smooth function, we will have
$\left\|hu_{1}\right\|_\infty \le \eta$
for $h$ small enough, and inequality (\ref{2eq:two}) will hold,
yielding thereby:
\begin{equation}
  \psi (hu_{1}) \le \frac{h^{2}}{2}
  \frac{1}{\lambda} \left\|u_{1} \right\|_{2}^{2} + \frac{h^{2}}{2}
  \frac{1}{\delta '} \left\|u_{1}\right\|^{2}\ .
\end{equation}

If we choose $\delta '$ close enough to $\delta$, the quantity
$\left(\frac{1}{\lambda} + \frac{1}{\delta '}\right)$
will be negative, and we end up with
\begin{equation}
  \psi (hu_{1}) < 0\ \ \ \ \ {\rm for}\ \ h\ne 0\ \ {\rm small}\ .
\end{equation}

On the other hand, we check directly that $\psi (0) = 0$. This shows
that 0 cannot be a minimizer of $\psi$, not even a local one.
So $\overline{u} \ne 0$ and
$\overline{u} \ne \Lambda_{o}^{-1} (0) = 0$. \qed
\end{proof}
%
\begin{corollary}
Assume $H$ is $C^{2}$ and
$\left(a_{\infty},b_{\infty}\right)$-subquadratic at infinity. Let
$\xi_{1},\allowbreak\dots,\allowbreak\xi_{N}$  be the
equilibria, that is, the solutions of $H' (\xi ) = 0$.
Denote by $\omega_{k}$
the smallest eigenvalue of $H'' \left(\xi_{k}\right)$, and set:
\begin{equation}
  \omega : = {\rm Min\,} \left\{\omega_{1},\dots,\omega_{k}\right\}\ .
\end{equation}
If:
\begin{equation}
  \frac{T}{2\pi} b_{\infty} <
  - E \left[- \frac{T}{2\pi}a_{\infty}\right] <
  \frac{T}{2\pi}\omega
  \label{2eq:three}
\end{equation}
then minimization of $\psi$ yields a non-constant $T$-periodic solution
$\overline{x}$.
\end{corollary}
%

We recall once more that by the integer part $E [\alpha ]$ of
$\alpha \in \bbbr$, we mean the $a\in \bbbz$
such that $a< \alpha \le a+1$. For instance,
if we take $a_{\infty} = 0$, Corollary 2 tells
us that $\overline{x}$ exists and is
non-constant provided that:

\begin{equation}
  \frac{T}{2\pi} b_{\infty} < 1 < \frac{T}{2\pi}
\end{equation}
or
\begin{equation}
  T\in \left(\frac{2\pi}{\omega},\frac{2\pi}{b_{\infty}}\right)\ .
  \label{2eq:four}
\end{equation}

%
\begin{proof}
The spectrum of $\Lambda$ is $\frac{2\pi}{T} \bbbz +a_{\infty}$. The
largest negative eigenvalue $\lambda$ is given by
$\frac{2\pi}{T}k_{o} +a_{\infty}$,
where
\begin{equation}
  \frac{2\pi}{T}k_{o} + a_{\infty} < 0
  \le \frac{2\pi}{T} (k_{o} +1) + a_{\infty}\ .
\end{equation}
Hence:
\begin{equation}
  k_{o} = E \left[- \frac{T}{2\pi} a_{\infty}\right] \ .
\end{equation}

The condition $\gamma < -\lambda < \delta$ now becomes:
\begin{equation}
  b_{\infty} - a_{\infty} <
  - \frac{2\pi}{T} k_{o} -a_{\infty} < \omega -a_{\infty}
\end{equation}
which is precisely condition (\ref{2eq:three}).\qed
\end{proof}
%

\begin{lemma}
Assume that $H$ is $C^{2}$ on $\bbbr^{2n} \setminus \{ 0\}$ and
that $H'' (x)$ is non-de\-gen\-er\-ate for any $x\ne 0$. Then any local
minimizer $\widetilde{x}$ of $\psi$ has minimal period $T$.
\end{lemma}
%
\begin{proof}
We know that $\widetilde{x}$, or
$\widetilde{x} + \xi$ for some constant $\xi
\in \bbbr^{2n}$, is a $T$-periodic solution of the Hamiltonian system:
\begin{equation}
  \dot{x} = JH' (x)\ .
\end{equation}

There is no loss of generality in taking $\xi = 0$. So
$\psi (x) \ge \psi (\widetilde{x} )$
for all $\widetilde{x}$ in some neighbourhood of $x$ in
$W^{1,2} \left(\bbbr / T\bbbz ; \bbbr^{2n}\right)$.

But this index is precisely the index
$i_{T} (\widetilde{x} )$ of the $T$-periodic
solution $\widetilde{x}$ over the interval
$(0,T)$, as defined in Sect.~2.6. So
\begin{equation}
  i_{T} (\widetilde{x} ) = 0\ .
  \label{2eq:five}
\end{equation}

Now if $\widetilde{x}$ has a lower period, $T/k$ say,
we would have, by Corollary 31:
\begin{equation}
  i_{T} (\widetilde{x} ) =
  i_{kT/k}(\widetilde{x} ) \ge
  ki_{T/k} (\widetilde{x} ) + k-1 \ge k-1 \ge 1\ .
\end{equation}

This would contradict (\ref{2eq:five}), and thus cannot happen.\qed
\end{proof}
%
\paragraph{Notes and Comments.}
The results in this section are a
refined version of \cite{2clar:eke};
the minimality result of Proposition
14 was the first of its kind.

To understand the nontriviality conditions, such as the one in formula
(\ref{2eq:four}), one may think of a one-parameter family
$x_{T}$, $T\in \left(2\pi\omega^{-1}, 2\pi b_{\infty}^{-1}\right)$
of periodic solutions, $x_{T} (0) = x_{T} (T)$,
with $x_{T}$ going away to infinity when $T\to 2\pi \omega^{-1}$,
which is the period of the linearized system at 0.

\begin{table}
\caption{This is the example table taken out of {\it The
\TeX{}book,} p.\,246}
\begin{center}
\begin{tabular}{r@{\quad}rl}
\hline
\multicolumn{1}{l}{\rule{0pt}{12pt}
                   Year}&\multicolumn{2}{l}{World population}\\[2pt]
\hline\rule{0pt}{12pt}
8000 B.C.  &     5,000,000& \\
  50 A.D.  &   200,000,000& \\
1650 A.D.  &   500,000,000& \\
1945 A.D.  & 2,300,000,000& \\
1980 A.D.  & 4,400,000,000& \\[2pt]
\hline
\end{tabular}
\end{center}
\end{table}
%
\begin{theorem} [Ghoussoub-Preiss]
Assume $H(t,x)$ is
$(0,\varepsilon )$-subquadratic at
infinity for all $\varepsilon > 0$, and $T$-periodic in $t$
\begin{equation}
  H (t,\cdot )\ \ \ \ \ {\rm is\ convex}\ \ \forall t
\end{equation}
\begin{equation}
  H (\cdot ,x)\ \ \ \ \ {\rm is}\ \ T{\rm -periodic}\ \ \forall x
\end{equation}
\begin{equation}
  H (t,x)\ge n\left(\left\|x\right\|\right)\ \ \ \ \
  {\rm with}\ \ n (s)s^{-1}\to \infty\ \ {\rm as}\ \ s\to \infty
\end{equation}
\begin{equation}
  \forall \varepsilon > 0\ ,\ \ \ \exists c\ :\
  H(t,x) \le \frac{\varepsilon}{2}\left\|x\right\|^{2} + c\ .
\end{equation}

Assume also that $H$ is $C^{2}$, and $H'' (t,x)$ is positive definite
everywhere. Then there is a sequence $x_{k}$, $k\in \bbbn$, of
$kT$-periodic solutions of the system
\begin{equation}
  \dot{x} = JH' (t,x)
\end{equation}
such that, for every $k\in \bbbn$, there is some $p_{o}\in\bbbn$ with:
\begin{equation}
  p\ge p_{o}\Rightarrow x_{pk} \ne x_{k}\ .
\end{equation}
\qed
\end{theorem}
%
\begin{example} [{{\rm External forcing}}]
Consider the system:
\begin{equation}
  \dot{x} = JH' (x) + f(t)
\end{equation}
where the Hamiltonian $H$ is
$\left(0,b_{\infty}\right)$-subquadratic, and the
forcing term is a distribution on the circle:
\begin{equation}
  f = \frac{d}{dt} F + f_{o}\ \ \ \ \
  {\rm with}\ \ F\in L^{2} \left(\bbbr / T\bbbz; \bbbr^{2n}\right)\ ,
\end{equation}
where $f_{o} : = T^{-1}\int_{o}^{T} f (t) dt$. For instance,
\begin{equation}
  f (t) = \sum_{k\in \bbbn} \delta_{k} \xi\ ,
\end{equation}
where $\delta_{k}$ is the Dirac mass at $t= k$ and
$\xi \in \bbbr^{2n}$ is a
constant, fits the prescription. This means that the system
$\dot{x} = JH' (x)$ is being excited by a
series of identical shocks at interval $T$.
\end{example}
%
\begin{definition}
Let $A_{\infty} (t)$ and $B_{\infty} (t)$ be symmetric
operators in $\bbbr^{2n}$, depending continuously on
$t\in [0,T]$, such that
$A_{\infty} (t) \le B_{\infty} (t)$ for all $t$.

A Borelian function
$H: [0,T]\times \bbbr^{2n} \to \bbbr$
is called
$\left(A_{\infty} ,B_{\infty}\right)$-{\it subquadratic at infinity}
if there exists a function $N(t,x)$ such that:
\begin{equation}
  H (t,x) = \frac{1}{2} \left(A_{\infty} (t) x,x\right) + N(t,x)
\end{equation}
\begin{equation}
  \forall t\ ,\ \ \ N(t,x)\ \ \ \ \
  {\rm is\ convex\ with\  respect\  to}\ \ x
\end{equation}
\begin{equation}
  N(t,x) \ge n\left(\left\|x\right\|\right)\ \ \ \ \
  {\rm with}\ \ n(s)s^{-1}\to +\infty\ \ {\rm as}\ \ s\to +\infty
\end{equation}
\begin{equation}
  \exists c\in \bbbr\ :\ \ \ H (t,x) \le
  \frac{1}{2} \left(B_{\infty} (t) x,x\right) + c\ \ \ \forall x\ .
\end{equation}

If $A_{\infty} (t) = a_{\infty} I$ and
$B_{\infty} (t) = b_{\infty} I$, with
$a_{\infty} \le b_{\infty} \in \bbbr$,
we shall say that $H$ is
$\left(a_{\infty},b_{\infty}\right)$-subquadratic
at infinity. As an example, the function
$\left\|x\right\|^{\alpha}$, with
$1\le \alpha < 2$, is $(0,\varepsilon )$-subquadratic at infinity
for every $\varepsilon > 0$. Similarly, the Hamiltonian
\begin{equation}
H (t,x) = \frac{1}{2} k \left\|k\right\|^{2} +\left\|x\right\|^{\alpha}
\end{equation}
is $(k,k+\varepsilon )$-subquadratic for every $\varepsilon > 0$.
Note that, if $k<0$, it is not convex.
\end{definition}
%

\paragraph{Notes and Comments.}
The first results on subharmonics were
obtained by Rabinowitz in \cite{2rab}, who showed the existence of
infinitely many subharmonics both in the subquadratic and superquadratic
case, with suitable growth conditions on $H'$. Again the duality
approach enabled Clarke and Ekeland in \cite{2clar:eke:2} to treat the
same problem in the convex-subquadratic case, with growth conditions on
$H$ only.

Recently, Michalek and Tarantello (see Michalek, R., Tarantello, G.
\cite{2mich:tar} and Tarantello, G. \cite{2tar}) have obtained lower
bound on the number of subharmonics of period $kT$, based on symmetry
considerations and on pinching estimates, as in Sect.~5.2 of this
article.

%
% ---- Bibliography ----
%
\begin{thebibliography}{}
%
\bibitem[1980]{2clar:eke}
Clarke, F., Ekeland, I.:
Nonlinear oscillations and
boundary-value problems for Hamiltonian systems.
Arch. Rat. Mech. Anal. {\bf 78} (1982) 315--333

\bibitem[1981]{2clar:eke:2}
Clarke, F., Ekeland, I.:
Solutions p\'{e}riodiques, du
p\'{e}riode donn\'{e}e, des \'{e}quations hamiltoniennes.
Note CRAS Paris {\bf 287} (1978) 1013--1015

\bibitem[1982]{2mich:tar}
Michalek, R., Tarantello, G.:
Subharmonic solutions with prescribed minimal
period for nonautonomous Hamiltonian systems.
J. Diff. Eq. {\bf 72} (1988) 28--55

\bibitem[1983]{2tar}
Tarantello, G.:
Subharmonic solutions for Hamiltonian
systems via a $\bbbz_{p}$ pseudoindex theory.
Annali di Matematica Pura (to appear)

\bibitem[1985]{2rab}
Rabinowitz, P.:
On subharmonic solutions of a Hamiltonian system.
Comm. Pure Appl. Math. {\bf 33} (1980) 609--633

\end{thebibliography}
\clearpage
\addtocmark[2]{Author Index} % additional numbered TOC entry
\renewcommand{\indexname}{Author Index}
\printindex
\clearpage
\addtocmark[2]{Subject Index} % additional numbered TOC entry
\markboth{Subject Index}{Subject Index}
\renewcommand{\indexname}{Subject Index}
\input{subjidx.ind}
\end{document}

### 2014-11-25 thesis-proposal/problem.tex
Updated proposal

@@
 Much progress has been made in the decades since Reynolds' statement above, but  across all language lineages,  new language and tool \emph{dialects} remain common vehicles for new ideas, both in research and practice.
-This suggests that library-based mechanisms are still not general enough to encompass all desirable modes of expression.
+This suggests that library-based mechanisms are still not general enough to encompass many desirable modes of expression.
 %Library implementations are not yet satisfying.
@@
 In this thesis, we will design extension mechanisms that handle these issues, permitting library-based expression of features that today would need to be built into a system by its central designers.
-More specifically, we show how to delegate control over certain aspects of the concrete syntax, external type system and editor services to \emph{static functions}, i.e.
+More specifically, we show how to delegate control over certain aspects of the concrete syntax, static semantics and editor services to \emph{static functions}, i.e.
 user-defined functions written in a \emph{static language}.
-By layering these mechanisms over a fixed typed internal language, constraining the static language appropriately and validating the code it generates, we will retain key  metatheoretic properties and arrive at powerful modular reasoning principles.
+By layering these mechanisms over a fixed typed internal language, constraining the static language appropriately and validating the code it is used to generate, we will retain key  metatheoretic properties and arrive at powerful modular reasoning principles.
 %By associating these static functions with type constructors, forming \emph{active type constructors},  we will further show how they can enjoy a usage profile essentially identical to that of built in features.
@@
 \subsection{Motivating Examples}\label{sec:examples}
-To make the use cases we seek to address more concrete, let us begin with a few simple examples that we will return to throughout this work.
+To make the use cases we seek to address more concrete, we begin with a few examples of system features that we wish to make expressible.
 
@@
 For example, consider a bioinformatics  application: the BisI restriction enzyme cuts DNA when it sees the motif \texttt{GC\textit{N}GC}, where \texttt{\textit{N}} is any base.
-We would want to express such patterns with syntax like this, using curly braces to splice one pattern into another and parentheses to delimit captured groups:
+We would want to express such patterns with syntax like the following, where we use curly braces to splice one pattern into another and parentheses to delimit captured groups:
 \begin{lstlisting}[numbers=none]
@@
 %The cognitive load of reading and writing patterns is low and patterns are parsed once at compile-time. 
-Malformed patterns would ideally result in intelligible \emph{compile-time} parse errors.
+Malformed patterns, e.g. \lstinline{<SURLGC(EURL>} must result in intelligible \emph{compile-time} parse errors no matter where they appear in a program.
 \item A \textbf{static semantics} that ensures that key invariants related to regular expressions are statically maintained. For example, 
 	\begin{enumerate}
-	\item At minimum, we would want that only patterns, or properly escaped strings, are spliced into a pattern literal like that above, to avoid splicing errors and injection attacks \cite{owasp2013, Bravenboer:2007:PIA:1289971.1289975}.
+	\item At minimum, we would want that only patterns, or properly escaped strings, are spliced into a pattern literal, to avoid splicing errors and injection attacks \cite{owasp2013, Bravenboer:2007:PIA:1289971.1289975}.
 For example,
@@
 %Unfortunately, it is impossible to completely define the syntax and the semantics referenced above using contemporary mechanisms, so library providers need to compromise.
+
 The most common strategy, exemplified by Java, is to ask clients to introduce patterns as strings, deferring their  parsing and compilation to run-time.
@@
 Parsing errors due to these clashes have been observed to be difficult for even experienced programmers to diagnose \cite{Omar:2012:ACC:2337223.2337324}.
-Moreover, the static guarantees cannot be provided, leading to run-time exceptions (common even in well-tested code using regular expressions \cite{spishak2012type}), and mistakes that are not caught even at run-time, some of which can lead to security vulnerabilities due to injection attacks \cite{owasp2013}.
-It also introduces performance overhead due to run-time parsing and compilation of patterns, and redundant well-formedness checks.
+Moreover, the static semantics described in feature 2 cannot be expressed, leading to run-time exceptions (even in well-tested code using regular expressions \cite{spishak2012type}), and mistakes that cannot be  caught even with run-time checks.
+This can lead to security vulnerabilities due to injection attacks, a widespread contemporary problem \cite{owasp2013}.
+It also introduces performance overhead due to run-time parsing and compilation of patterns.
 
@@
 \ref{sec:syntax}, or a class hierarchy in an object-oriented language), or using an abstract data type.
-This provides feature 2a while sacrificing feature 1 (such an encoding affords only an approximation of the verbose abstract syntax of patterns, not their concrete syntax).
-Advanced invariants like those described in features 2b are not directly tracked by such encodings and require solutions beyond those commonly found in simply-typed languages (a feature like GADTs might suffice in this case \cite{XiCheChe03}).%Explicitly constructing a regular expression by applying datatype constructors is unacceptably verbose.
+This provides feature 2a at the expense of feature 1 (such an encoding affords only an approximation of the verbose abstract syntax of patterns, not the concrete syntax above).
+Advanced invariants like those described in feature 2b are not directly tracked by the simplest such encodings (languages with support for GADTs could be used in this particular case \cite{XiCheChe03}).%Explicitly constructing a regular expression by applying datatype constructors is unacceptably verbose.
 
 None of these approaches address the important issue of tool support (feature 3).
-A number of tools are available online that provide services ranging from simple syntax highlighting to interactive testing and explanations of complex patterns \cite{regexr} and pattern extraction from examples (e.g.
-\cite{_txt2re:_????}).
+A number of tools are available online that provide services ranging from simple syntax highlighting to interactive testing and explanations of complex patterns \cite{regexr} and pattern extraction from examples \cite{_txt2re:_????}.
 We have found that programmers benefit substantially from their use \cite{Omar:2012:ACC:2337223.2337324}.
@@
 \subsubsection{Example 2: Regular Strings}\label{sec:rstr}
-The examples above all dealt with how regular expression patterns are represented.
+The example above dealt with how regular expression patterns are represented.
 However, programmers also want to reason about which regular language a string value is in \cite{sanitation-psp14}.
-For example, we might want to ensure that the arguments to a function called \verb|connect| corresponding to a database username and password are only alphanumeric strings.
-We might include this information directly in the type of \verb|connect|, i.e.
+For example, we might want to ensure that the arguments to a function that creates a database connection given a username and password are only alphanumeric strings.
+We would ideally include this information directly in the type of the  function:
 \begin{lstlisting}[numbers=none]
@@
 
-This example uses a specialized \emph{type constructor}, \verb|rstring|, indexed by a statically known regular expression.
-The operations on such \emph{regular strings} would need a static semantics that track the regular language a string is in.
-For example, a regular string introduced by concatenating two other alphanumeric strings  should also be an alphanumeric string.
-
-Ideally, we would be able to use standard string literal syntax with regular strings, e.g. 
+This example assumes that the language includes a \emph{type constructor}, \verb|rstring|, indexed by a statically known regular expression, written using the syntax described above.
+Ideally, we would be able to use standard string literal syntax for such regular strings:
 \begin{lstlisting}[numbers=none]
@@
 
-Regular strings could go beyond simply refining standard string types. For example, they might define operators specific to regular strings like \emph{captured group projection}:
+To be useful, the language would need to come equipped with a static semantics that continues to track the regular language a string is in through standard string operations.
+For example, a regular string introduced by concatenating two other alphanumeric strings  should also be an alphanumeric string.
+All such operations should have the same or lower cost when compared to the corresponding operations on standard strings.
+
+Regular strings might go beyond simply \emph{refining} standard string types (i.e.
+specifying verification conditions atop an existing semantics) \cite{Freeman91}.
+For example, they might define an operation like \emph{captured group projection} that cannot be defined for strings:
 \begin{lstlisting}[numbers=none]
@@
 \end{lstlisting}
-Ideally, it would be possible to define this operator such that its cost is $\mathcal{O}(1)$.
+It should be possible to give this operation a cost of $\mathcal{O}(1)$.
+From a language implementation perspective, this requires that different regular string types  have different representations, also not possible when using a refinement system over strings.
+%This is not possible if regular strings are defined using a refinement system atop a language with only strings.
 %This example requires introducing new operators into the system that did not already exist, and using an underlying represention that differs for different type indices.
@@
 \subsubsection{Example 3: Labeled Products with Functional Update Operators}\label{sec:lprod}
-The simplest semantics for product types is to define only nullary and binary products \cite{pfpl}.
+The simplest way to specify the semantics of product types is to define only nullary and binary products \cite{pfpl}.
 However, in practice, many variations on product types are built in to various dialects of ML and other languages: $n$-ary tuples, labeled tuples,
 records (identified up to reordering), and 
-records with width and depth coercions \cite{Cardelli:1984:SMI:1096.1098} and functional update operators \cite{ocaml-manual}.
-{The Haskell wiki notes that ''No, extensible records [i.e.
-records with a functional update operator] are not implemented in GHC.
+records with width and depth coercions \cite{Cardelli:1984:SMI:1096.1098} and functional update operators (also called \emph{extensible records}) \cite{ocaml-manual}.
+{The Haskell wiki notes that ''extensible records  are not implemented in GHC.
 The problem is that the record design space is large, and seems to lack local optima.
@@
 
-We would ideally like to avoid needing the language designer to decide \emph{a priori} on just one point in this large design space.
-Instead, they should simply provide some minimal way of defining products, perhaps only nullary and binary products, while making it possible for these other variations on products to be defined as libraries and used together without conflict.
-For example, we would like to be able to define labeled product types, which are like record types but maintain a row ordering.
-An example of a labeled product types classifying conference papers might be:
+We would ideally like to avoid needing the language designer to decide \emph{a priori} on just one privileged point in this large design space.
+Instead, they should simply provide some minimal mechanism for defining products of values, e.g.
+only nullary and binary products, and include a mechanism that makes it possible for these other variations on products to be defined as libraries and then used together without conflict.
+For example, we would like to define the semantics of labeled tuples, which have labeled rows like records but maintain a row ordering like tuples, in a library.
+An example of a labeled product type classifying conference papers might be:
 \begin{lstlisting}[numbers=none]
@@
 
-Because of the row ordering, it should be possible to introduce values of this type with or without explicit labels, e.g.
+The row ordering should make it possible to introduce values of this type with or without explicit labels, e.g.
 \begin{lstlisting}[numbers=none]
@@
 \end{lstlisting}
-Note above that we aim to be able to re-use the standard record and tuple syntax.
+Note that we aim to be able to unambiguously re-use standard record and tuple syntax.
 
-We should then be able to project out rows by providing a label (or a numeric index, not shown):
+We should then be able to project out rows by providing a label:
 \begin{lstlisting}[numbers=none]
@@
 
-We might also be wish to support functional update. For example, an operation that dropped a row might be used like this:
+We might also wish for our labeled tuples to support functional update. For example, an operation that dropped a row might be used like this:
 \begin{lstlisting}[numbers=none]
@@
 \end{lstlisting}
-An operation that added a row, or updated the value of an existing row, might be used like this:
+An operation that added a row, or updated an existing row, might be used like this:
 \begin{lstlisting}[numbers=none]
-fun with_author(p : Paper, a : rstring[<.+>]) = p.ext(author=a)
+fun with_author(p : Paper, a : rstring[<SURL.+EURL>]) = p.ext(author=a)
 \end{lstlisting}
@@
 
-In situations, like these, where a library-based approach is not satisfying, providers must today take a \emph{language-external approach}, either by specifying a new system dialect (implementing it using a \emph{compiler generator} \cite{brooker1963compiler}, \emph{language workbench} \cite{erdweg2013state}, \emph{DSL framework} \cite{fowler2010domain}, or simply by forking an existing codebase), or by using an extension mechanism for a {particular} compiler\footnote{Compilers that modify, or allow modification of, the semantics of their base language, rather than simply permitting semantics-preserving optimizations, should be considered a pernicious means for creating new language dialects.
-That is, some programs that purport to be written in C, Haskell or Standard ML are actually written in compiler-specific dialects of these languages.}, editor or other tool.
-For example, a researcher interested in providing the regular expression related features just described (let us refer to these collectively as \texttt{R}) might design a new system with built-in support for them, perhaps basing it on an existing system containing some general-purpose features (\texttt{G}).
-A different researcher developing a new language-integrated parallel programming abstraction (\texttt{P}) might  take the same approach.
-A third researcher, developing a type system for reasoning about units of measure (\texttt{Q}) might again do the same.
-This results in a collection of distinct programming systems, as diagrammed in Figure \ref{approaches}a.
+In situations like these where a library-based approach is not satisfying, providers must today take a \emph{language-external approach}, specifying a new system dialect and implementing it using a \emph{compiler generator} \cite{brooker1963compiler}, \emph{language workbench} \cite{erdweg2013state}, \emph{DSL framework} \cite{fowler2010domain}, using an extension mechanism for a {particular} compiler\footnote{Compilers that modify, or allow modification of, the semantics of their base language, rather than simply permitting semantics-preserving optimizations, should be considered a pernicious means for creating new language dialects.
+%That is, some programs that purport to be written in C, Haskell or Standard ML are actually written in compiler-specific dialects of these languages.
+}, editor or other tool, or by forking an existing implementation.
+For example, a researcher interested in providing the regular expression related features just described (let us refer to these collectively as \texttt{R}) might design a new system with built-in support for them, perhaps basing it on an existing system containing some general-purpose features (\texttt{G}, e.g.
+the features of Standard ML).
+A different researcher developing a new language-integrated parallel programming abstraction (\texttt{P}) might  take the same approach (e.g.
+Manticore is such a system \cite{conf/popl/FluetRRSX07}).
+A third researcher, developing a type system for reasoning about units of measure (\texttt{Q}) might again do the same \cite{conf/cefp/Kennedy09}.
+This results in a collection of distinct  systems, as diagrammed in Figure \ref{approaches}a.
 
@@
 That is, one must either use the system containing features \texttt{G+R}, \texttt{G+P} or \texttt{G+Q}.
-There is no system containing \texttt{G}, \texttt{R}, \texttt{P} and \texttt{Q} in other combinations, and merging the systems containing each separately can be non-trivial because there can be serious interference and safety issues, as we will discuss at length throughout this thesis.
+There is no system containing \texttt{G}, \texttt{R}, \texttt{P} and \texttt{Q} in other combinations, and merging the systems defining each separately can be non-trivial because there can be serious interference and safety issues, as we will discuss at length throughout this thesis.
 % More specifically, there is no guarantee of \textbf{orthogonality} or even \textbf{interoperability}.% This has limited the broad adoption of these kinds of innovations.%This latter method couples the semantics of the feature to the implementation details of a particular tool.
@@
 
-Recent evidence indicates that this is one of the major barriers preventing research  from being driven into practice.
-For example, developers prefer high-level language-integrated parallel programming abstractions that provide  stronger semantic guarantees  \cite{cave2010comparing}, but library-based approximations are far more widely adopted because  ``parallel programming languages'' privilege only a few  abstractions at the language level.
-In contrast, it is widely acknowledged that  different abstractions are more appropriate in different situations \cite{Tasharofi:2013rc}.
-Moreover,  parallel programming is rarely the only relevant concern outside of a classroom or research setting.
-Support for regular expressions as above would be simultaneously desirable for processing large amounts of genomic data.% but using these features together in the same compilation unit would be difficult or impossible if implemented using language-external means.
+% Recent evidence indicates that this is one of the major barriers preventing research  from being driven into practice.
+For example, developers prefer high-level language-integrated parallel programming abstractions that provide  stronger semantic guarantees  \cite{cave2010comparing}, but library-based approximations are far more widely adopted because the ``parallel programming languages'' privilege only a few  specialized abstractions at the language level.
+In contrast, it is widely acknowledged that  different specialized abstractions are more appropriate in different situations \cite{Tasharofi:2013rc}.
+Moreover,  parallel programming is rarely the only relevant specialized concern.
+Support for regular expressions as above would be simultaneously desirable for processing large amounts of genomic data in parallel.% but using these features together in the same compilation unit would be difficult or impossible if implemented using language-external means.
 Indeed, switching to a ``parallel programming language'' would likely make it \emph{more} difficult to use regular expressions, as these are likely to be less well-developed in a specialized language than in an established general-purpose language.% This intuition was perhaps most succinctly expressed by a participant in a recent study by Basili et al.
@@
 \caption{\small (a) When taking a language-external approach, new features are packaged together into separate languages and tools.
-(b) When taking a language-integrated approach, there is one extensible host language and the compile-time and edit-time logic governing new constructs is expressed within ``active libraries''.}
+(b) We propose an approach where there is one extensible host language and the compile-time and edit-time logic governing new constructs is expressed within safely composable libraries.}
 \label{approaches}
@@
 \end{figure}
-We argue that, due to these problems, taking a language-external approach to realizing a new feature should be considered harmful and avoided whenever possible.
-The goal of the research being proposed here is to design \emph{language-integrated extension mechanisms} that give providers the ability to define, within libraries, new features that, like those above, that have previously required central planning,
+We argue that, due to these problems of composition, taking a language-external approach to realizing a new feature should be considered harmful and avoided whenever possible.
+The goal of the research being proposed here is to design \emph{language-integrated extension mechanisms} that give providers the ability to define, within libraries, new features like those above that have previously required central planning,
 %\footnote{One might compare today's programming systems to  {centrally-planned} economies, whereas extensible\- systems more closely resemble modern market economies.
@@
 We leave further development of this analogy to the reader.}
-so that language-external approaches are less frequently necessary, as illustrated in Figure \ref{approaches}b. 
-Such libraries have been called \emph{active libraries}  \cite{activelibraries} in that they more actively influence the syntax and semantics of the system.
+so that language-external approaches are less frequently necessary. The concept is  illustrated in Figure \ref{approaches}b. %
+%Such libraries have been called \emph{active libraries}  \cite{activelibraries}.
 %Features implemented within active libraries can be imported individually, unlike features implemented by external means, giving us a potential means to avoid the problems of orthogonality and interoperability just described.
@@
 We must proceed with caution, however: critical issues having to do with {safety} must be overcome before language-integrated extension mechanisms can be introduced into a system.
-If too much control over  these core features of the system is given  to developers, the system may become quite unreliable.
+If too much control over  these core features of the system is given  to developers, the system's metatheory may become weak.
 %For example, an extension could weaken important metatheoretic guarantees previously provided by the system. 
-Type safety, for example, may not hold if the static and dynamic semantics of the language can be modified or extended arbitrarily from within libraries.
-Furthermore, even if extensions can be shown not to cause such problems in isolation, there may still be conflicts between extensions that could weaken their semantics, leading to subtle problems that only appear when two extensions are used together, thwarting any attempts to reason modularly about particular programs and the system as a whole.
+For example, parsing ambiguities might arise if the syntax can be modified arbitrarily.
+Type safety  may not hold if the static and dynamic semantics of the language can be modified or extended arbitrarily from within libraries.
+Furthermore, even if extensions can be shown not to cause such problems in isolation, there may still be conflicts between extensions that could weaken their semantics, leading to subtle problems that only appear when two extensions are used together.
 %As a simple example, if two active libraries introduce the same syntactic form but back it with differing (but individually valid) semantics, this ambiguity  would only manifest itself when both libraries were imported within the same scope.

### 2014-12-02 sanitation-psp14/final.tex
prep for camera ready.
Edits for camera ready.

@@
-\documentclass[9pt]{sig-alternate}
+\documentclass[10pt]{sigplanconf}
 \newcommand{\lamAce}{\lambda_{\text{Ace}}}
@@
 \usepackage{amssymb} 
+\def \TirNameStyle #1{\footnotesize \textsc{#1}}
+\renewcommand{\MathparLineskip}{\lineskiplimit=.4\baselineskip\lineskip=.4\baselineskip plus .2\baselineskip}
 
@@
 \usepackage{amsthm}
+\usepackage{thmtools}
 
@@
 
-\usepackage{hyperref}		% For links in citations
+\usepackage[hidelinks]{hyperref}		% For links in citations
 
@@
 \newtheorem{thm}{Theorem}                                                       
+\declaretheoremstyle[%
+  spaceabove=-6pt,%
+  spacebelow=6pt,%
+  headfont=\normalfont\itshape,%
+  postheadspace=1em,%
+  qed=\qedsymbol%
+]{mystyle} 
+\declaretheorem[name={Proof Sketch},style=mystyle,unnumbered
+]{prf}
 \newtheorem{cor}[thm]{Corollary}                                                
@@
 \newtheorem{rem}[thm]{Remark} 
-\renewcommand*{\proofname}{Proof Sketch}
+%\renewcommand*{\proofname}{Proof Sketch}
 
@@
 \newcommand{\inferlbl}[3]
-			{\inferrule{#3}{#2}{\textsf{\footnotesize{\sc #1}}}}
+			{\inferrule{#3}{#2}{\textsf{\scriptsize{\sc #1}}}}
 \newcommand{\inferline}[3]
-			{\inferrule{#3}{#2} & {\textsf{\footnotesize{\sc #1}}} \\ \\}
+			{\inferrule{#3}{#2} & {\textsf{\scriptsize{\sc #1}}} \\ \\}
 
@@
 % target language terms.
-\newcommand{\tsubst}[3]{{\sf replace}(#1;#2;#3)} \newcommand{\metareplace}[3]{{\sf replace}(#1;#2;#3)} % TODO-nrf rename the commands. Lots of legacy naming around these parts...
-
-\newcommand{\tcheck}[4]{{\sf check}(#1; #2; #3; #4)}
+\newcommand{\tsubst}[3]{{\sf preplace}(#1;#2;#3)} \newcommand{\metareplace}[3]{{\sf replace}(#1;#2;#3)} % TODO-nrf rename the commands. Lots of legacy naming around these parts...
+
+\newcommand{\tcheck}[4]{{\sf pcheck}(#1; #2; #3; #4)}
 \renewcommand{\tstr}[1]{{{\sf str}[#1]}}
-\newcommand{\preplace}[3]{{\sf replace}(#1;#2;#3)}
-\newcommand{\tconcat}[2]{{\sf concat}(#1;#2)} \newcommand{\concat}[2]{{\sf concat}(#1;#2)} % lots of legacy naming around these parts...
+\newcommand{\preplace}[3]{{\sf preplace}(#1;#2;#3)}
+\newcommand{\tconcat}[2]{{\sf pconcat}(#1;#2)} \newcommand{\concat}[2]{{\sf concat}(#1;#2)} % lots of legacy naming around these parts...
 
@@
 % Meta-theoretic functions
-\newcommand{\lsubst}[3]{{\sf subst}(#1;#2;#3)} % This used to renderlreplace(...) so there're probably mistkes wherever this command was used now.
+\newcommand{\lsubst}[3]{{\sf replace}(#1;#2;#3)} % This used to renderlreplace(...) so there're probably mistkes wherever this command was used now.
 \newcommand{\lreplace}[3]{{\sf lreplace}(#1; #2; #3)}
@@
 \newcommand{\strcase}[3]{ {\sf rstrcase}(#1; #2; #3)}
-\newcommand{\pstrcase}[3]{ {\sf strcase}(#1; #2; #3)}
+\newcommand{\pstrcase}[3]{ {\sf pstrcase}(#1; #2; #3)}
 
@@
 
+\usepackage[compact]{titlesec}
+\titlespacing{\section}{0pt}{6pt}{2pt}
+\titlespacing{\subsection}{0pt}{3pt}{0pt}
+\titlespacing{\subsubsection}{0pt}{3pt}{0pt}
+\titlespacing{\paragraph}{0pt}{3pt}{5pt}
+
 \begin{document}
@@
 
+\permissiontopublish
+\conferenceinfo{PSP~'14}{October 21, 2014, Portland, OR, USA.}
+\copyrightyear{2014}
+\copyrightdata{978-1-4503-2296-6/14/10}
+\doi{2687148.2687152} 
 \title{Statically Typed String Sanitation Inside a Python}
-\numberofauthors{3}
-\author{
-  \alignauthor
-  Nathan Fulton
-  \alignauthor
-  Cyrus Omar
-  \alignauthor
-  Jonathan Aldrich
-  \and
-  \affaddr{Carnegie Mellon University, Pittsburgh, PA}\\
-  %\affaddr{Pittsburgh, PA}\\
-  \email{\{nathanfu, comar, aldrich\}@cs.cmu.edu}
-}
-%\authorinfo{~}{~}{~}
-%\authorinfo{Nathan Fulton\and Cyrus Omar\and Jonathan Aldrich}
- %          {School of Computer Science\\
-  %          Carnegie Mellon University
-   %        \{nathanfu, comar, aldrich\}@cs.cmu.edu}
+%\numberofauthors{3}
+%\author{
+%  \alignauthor
+%  Nathan Fulton
+%  \alignauthor
+%  Cyrus Omar
+%  \alignauthor
+%  Jonathan Aldrich
+%  \and
+%  \affaddr{Carnegie Mellon University, Pittsburgh, PA}\\
+%  %\affaddr{Pittsburgh, PA}\\
+%  \email{\{nathanfu, comar, aldrich\}@cs.cmu.edu}
+%}
+%%\authorinfo{~}{~}{~}
+\authorinfo{Nathan Fulton\and Cyrus Omar\and Jonathan Aldrich}
+          {Carnegie Mellon University \\ 
+        \{nathanfu, comar, aldrich\}@cs.cmu.edu}
+%
 
@@
 
-%\category{D.3.2}{Programming Languages}{Language Classifications}[Extensible Languages]
-%\category{D.3.4}{Programming Languages}{Processors}[Compilers]
-%\category{F.3.1}{Logics \& Meanings of Programs}{Specifying and Verifying and Reasoning about Programs}[Specification Techniques]
-%\keywords
-%type-level computation, typed compilation
+%\category{D.3.2}{Programming Languages}{Language Classifications}[Extensible languages]
+\category{F.3.3}{Logics \& Meanings of Programs}{Studies of Program Constructs}[Type structure]
+%\category{F.4.3}{Logics \& Meanings of Programs}{Formal Languages}
+\keywords
+type systems, regular expressions, input sanitation, string sanitation, extensible languages, web security
+
 \section{Introduction}\label{intro}
@@
 \begin{figure}[t]
-\small
+\scriptsize
   \begin{grammar}
@@
 \begin{figure}[t]
-\small
+\scriptsize
   \begin{grammar}
 <$\sigma$> ::=  $\sigma \rightarrow \sigma$ | $\stringin{r}$    \hfill  source types
+\vspace{-8px}
+
+<$v$> ::= $\lambda x . e$ | $\strin{s}$ \hfill source values ($s \in \Sigma^{*}$) 
+\vspace{-8px}
 
 <$e$> ::= 
-      $x$ | $\lambda x . e$ | $e(e) $\hfill source terms \alt 
-      $\strin{s}$ | $\rsconcat{e}{e}$ | $\strcase{e}{e}{x,y.e}$ \hfill $s \in \Sigma^{*}$ \alt
+      $v$ | $x$ | $e(e)$   \hfill source terms \alt 
+      $\rsconcat{e}{e}$ | $\strcase{e}{e}{x,y.e}$ \alt
       $\rcoerce{r}{e}$ | $\rcheck{r}{e}{x.e}{e}$ | $\sisubst{r}{e}{e}$  
-
-<$v$> ::= $\lambda x . e$ | $\strin{s}$ \hfill source values 
       \caption{Syntax of $\lambda_{RS}$.}
@@
 
+\subsubsection{Concatenation and String Decomposition}
+
+The \textsc{S-T-Concat} rule is the simplest example of our approach.
+The rule is sound because the result of concatenating two strings, the first in $\lang{r_1}$ and the second in $\lang{r_2}$, will always be in the language $\lang{r_1\cdot r_2}$.
+The rule therefore relates string concatenation to sequential composition of regular expressions.
+
+Whereas concatenation allows the construction of large strings from smaller strings,  $\strcase{e}{e_0}{x,y.e_1}$  allows the decomposition, or elimination, of large strings into smaller strings.
+Intuitively, this operation branches based on whether a string is empty or not, exactly analagous to case analysis on lists in a functional language.
+The branch for a non-empty string ``peels off" the first character, binding it and the remainder of the string to specified variables.
+The evaluation rules \textsc{S-E-Case-$\epsilon$} and \textsc{S-E-Case-Concat} express this semantics.
+This construct can be used to implement a standard string indexing operation as a function, given a suitable definition of natural numbers (omitted).
+
+The regular expression recognizing any one-character prefix of the strings in $\lang{r}$ is easily defined.
+
 \begin{figure}[t]
-\small
+\scriptsize
 $\fbox{\inferrule{}{\sctx \vdash e : \sigma}}$
@@
 \inferrule[S-T-Replace]
-{\sctx \vdash e_1 : \stringin{r_1} \\ \sctx \vdash e_2 : \stringin{r_2} \\\\ \lreplace{r}{r_1}{r_2} = r'}
-{\sctx \vdash \sisubst{r}{e_1}{e_2} : \stringin{r'}}
+{ \sctx \vdash e_1 : \stringin{r_1}} 
+{\sctx \vdash \sisubst{r}{e_1}{e_2} : \stringin{ \lreplace{r}{r_1}{r_2} }}
 \end{mathpar}
-
+\vspace{-10px}
 \caption{Typing rules for $\lambdas$. 
@@
 
+\begin{defn}[Definition of $\lhead{r}$]
+$\lhead{r}$ is defined in terms of the structure of $r$:
+{\footnotesize
+\begin{align*}
+\lhead{r}                 &= \lhead{r, \epsilon} \\                             
+\lhead{\epsilon, r'}      &= \epsilon \\                                        
+\lhead{a, r'}             &= a \\                                               
+\lhead{r_1\cdot r_2, r'}  &= \lhead{r_1, r_2} \\                                
+\lhead{r_1 + r_2, r'}     &= \lhead{r_1, r'} + \lhead{r_2, r'} \\               
+\lhead{r^*, r'}           &= \lhead{r', \epsilon} + \lhead{r, \epsilon}         
+\end{align*}} 
+\end{defn}
+We use a two-argument auxiliary $\lhead{r, r'}$ in the definition of $\lhead{r}$ because when $r = q^*\cdot t$,
+$\lhead{r}$ needs to ``remember" $t$ in case $q$ is iterated zero times.
+
+Given this definition of $\lhead{r}$, regular expression derivatives \cite{bowzer} provide a useful tool for defining $\ltail{r}$.
+\begin{defn}[Brzozowski's Derivative]\label{def:derivative}
+  The \emph{derivative of $r$ with respect to $s$} 
+  is $\delta_s(r) = \{t | st \in \lang{r}\}$.
+\end{defn}
+
+Definition~\ref{def:derivative} is equivalent to the definition given in \cite{bowzer},
+although we refer the unfamiliar reader to \cite{owens}.
+Definition~\ref{def:derivative} is equivalent to Definition 3.1 in both papers.
+We now define $\ltail{r}$ using derivatives with respect to $\lhead{r}$.
+
+\begin{defn}[Definition of $\ltail{r}$]
+$\ltail{r}$ is defined in terms of $\lhead{r}$.
+Note that $\lhead{r} = a_1 + a_2 + ... + a_i$.
+We define $\ltail{r} = \delta_{a_1}(r) + \delta_{a_2}(r) + ... + \delta_{a_i}(r) + \epsilon$.
+\end{defn}
+
+The \textsc{S-T-Case} rule, which is defined in terms of these operations, thus relates the result of ``peeling off" the first character of a string to regular expression derivatives.
+
 \begin{figure}
-  \small
+  \scriptsize
 $\fbox{\inferrule{}{e \sreduces v}}$
@@
 \inferrule[S-E-Replace]
-{e_1 \sreduces \strin{s_1} \\ e_2 \sreduces \strin{s_2} \\ \lsubst{r}{s_1}{s_2} = s} 
-{\sisubst{r}{e_1}{e_2} \sreduces \sistr{s}}
+{e_1 \sreduces \strin{s_1} \\ e_2 \sreduces \strin{s_2}} 
+{\sisubst{r}{e_1}{e_2} \sreduces \sistr{ \lsubst{r}{s_1}{s_2} }}
 \end{mathpar}
+\vspace{-10px}
 \caption{Big step semantics for $\lambdas$}.
@@
 
-\subsubsection{Concatenation and String Decomposition}
-
-The \textsc{S-T-Concat} rule is the simplest example of our approach.
-The rule is sound because the result of concatenating two strings, the first in $\lang{r_1}$ and the second in $\lang{r_2}$, will always be in the language $\lang{r_1\cdot r_2}$.
-The rule therefore relates string concatenation to sequential composition of regular expressions.
-
-Whereas concatenation allows the construction of large strings from smaller strings,  $\strcase{e}{e_0}{x,y.e_1}$  allows the decomposition, or elimination, of large strings into smaller strings.
-Intuitively, this operation branches based on whether a string is empty or not, exactly analagous to case analysis on lists in a functional language.
-The branch for a non-empty string ``peels off" the first character, binding it and the remainder of the string to specified variables.
-The evaluation rules \textsc{S-E-Case-$\epsilon$} and \textsc{S-E-Case-Concat} express this semantics.
-This construct can be used to implement a standard string indexing operation as a function, given a suitable definition of natural numbers (omitted).
-
-The regular expressions describing the first character and remaining characters of a matching string in the language $\lang{r}$ must be determined to correctly specify the static semantics.
-For instance, the first character of a string in the language $\lang{(A\cdot B\cdot C)+(D\cdot E\cdot F)}$ is in the language $\lang{A+D}$ and the remaining characters are in the language $\lang{(B\cdot C)+(E \cdot F)}$.
-
-The regular expression recognizing any one-character prefix of the strings in $\lang{r}$ is easily defined.
-
-\begin{defn}[Definition of \lhead{r}]
-The relation $\lhead{r}=r'$ is defined in terms of the structure of $r$:
-\begin{itemize}[noitemsep]
-\item $\lhead{\epsilon} = \epsilon$
-\item $\lhead{.} = a_1 + a_2 + ... + a_n$ for all $a_i \in \Sigma$ where $|\Sigma| = n$.
-\item $\lhead{a} = a$ %where $a \in \Sigma$
-\item $\lhead{r_1\cdot r_2} = \lhead{r_1}$
-\item $\lhead{r_1 +  r_2}  = \lhead{r_1} + \lhead{r_2}$
-\item $\lhead{r*} = \epsilon + \lhead{r}$
-\end{itemize}
-\end{defn}
-
-Given this definition of $\lhead{r}$, regular expression derivatives \cite{bowzer} provide a useful tool for defining $\ltail{r}$.
-\begin{defn}[Brzozowski's Derivative]\label{def:derivative}
-  The \emph{derivative of $r$ with respect to $s$} is denoted by $\delta_s(r)$
-  and is $\delta_s(r) = \{t | st \in \lang{r}\}$.
-\end{defn}
-
-Definition~\ref{def:derivative} is equivalent to the definition given in \cite{bowzer},
-although we refer the unfamiliar reader to \cite{owens}.
-Definition~\ref{def:derivative} is equivalent to Definition 3.1 in both papers.
-We now define $\ltail{r}$ using derivatives with respect to $\lhead{r}$.
-
-\begin{defn}[Definition of \ltail{r}]
-The relation $\ltail{r} = r'$ is defined in terms of $\lhead{r}$.
-Note that $\lhead{r} = a_1 + a_2 + ... + a_i$.
-We define $\ltail{r} = \delta_{a_1}(r) + \delta_{a_2}(r) + ... + \delta_{a_i}(r)$.
-\end{defn}
-
-The \textsc{S-T-Case} rule, which is defined in terms of these operations, thus relates the result of ``peeling off" the first character of a string to regular expression derivatives.
-
 \subsubsection{Coercion}
@@
 % This form of coercion is useful because the replacement operation (introduced below in Section~\ref{sec:replace}) is often more conservative than absolutely necessary.
-Conversely, checked coercions, written $\rcheck{r}{e_0}{x.e_1}{e_2}$, 
+Conversely, checked coercions, $\rcheck{r}{e_0}{x.e_1}{e_2}$, 
 allow for passage to regular languages that are not necessarily smaller. Checked coercions branch based on whether the coercion succeeded or not.
@@
  \textsc{S-T-SafeCoerce} checks for language inclusion, which we write $\lang{r_1} \subseteq \lang{r_2}$.
-Language inclusion is a decidable relation.
+Language inclusion is decidable.
 As a result, the rule \textsc{S-E-SafeCoerce} does not need to perform any checks.
@@
 For safe coercions, it is often useful for the coercion to be performed implicitly.
-This can be seen as a form of  subtyping for regular string types \cite{fulton13}.
+This can be seen as a form of subtyping for regular string types \cite{gunter, fulton13, xhaskell2}.
 In practice, subtyping is crucial to the usability of our system due to the nature of the replacement operator.
-For simplicity, we do not include this relation; however, subtyping for regular strings is present in previous treatments of our work \cite{fulton12,fulton13}.
+For simplicity, we do not include this operator; however, subtyping for regular strings is present in previous treatments of our work \cite{fulton13}.
 The implementation discussed in Section \ref{atlang} also provides subtyping between regular string types.
@@
 It behaves analogously to \lstinline{str_replace} in PHP and \lstinline{String.replace} in Java, differing in that the replacement pattern $r$ must be statically given.
-The evaluation rule \textsc{S-E-replace} is defined via the relation $\lsubst{r}{s_1}{s_2}=s$.
-For simplicity, we do not inductively define this standard relation.
+The evaluation rule \textsc{S-E-replace} is defined in terms of $\lsubst{r}{s_1}{s_2}$.
+For simplicity, we do not define replacement in this paper; see \cite{psptr}.
 
 \begin{defn}[$\tt{subst}$]
-The relation $\lsubst{r}{s_1}{s_2} = s$ produces a string $s$ in which all substrings of $s_1$ in $\lang{r}$ are replaced with $s_2$.
+$\lsubst{r}{s_1}{s_2} = s$ produces a string $s$ in which all substrings of $s_1$ in $\lang{r}$ are replaced with $s_2$.
 \end{defn}
 
-The typing rule \textsc{S-T-Replace} requires computing the regular language of the string resulting from this operation given the languages of $s_1$ and $s_2$, written as the relation $\lreplace{r}{r_1}{r_2}={r'}$.
+The typing rule \textsc{S-T-Replace} requires computing the regular language of the string resulting from this operation given the languages of $s_1$ and $s_2$, written as $\lreplace{r}{r_1}{r_2}={r'}$.
 Given an automata-oriented interpretation of regular languages, it may be helpful to think in terms of replacing sub-automata.
@@
 A complete definition of ${\sf lreplace}$ would also give an rewrite system with correctness and termination proofs, which is beyond the scope of this paper.
-Instead, we provide an abstract definition of the relation and state necessary properties.
+Instead, we provide an abstract definition of the operation and state necessary properties.
 
 \begin{defn}[$\tt{lreplace}$]
-  The relation $\lreplace{r}{r_1}{r_2} = r'$ relates $r, r_1,$ and $r_2$ to
+  $\lreplace{r}{r_1}{r_2}$ relates $r, r_1,$ and $r_2$ to
   a language $r'$ containing all strings of $r_1$ except that any substring $s_{pre} s s_{post} \in \lang{r_1}$ where $s \in \lang{r}$
@@
 \end{prop}
-\begin{proof}
+\begin{prf}
 Algorithms for the inclusion problem may be adopted to identify any sublanguage $x \subseteq r$ of $r_1$.
@@
 Then $r'$ is $x_{pre}r_2x_{post}$.
-\end{proof}
-
-\begin{prop}[Substitution Correspondence] \label{thm:substcorrespondence}
-  If $s_1 \in \lang{r_1}$ and $s_2 \in \lang{r_2}$ then $\lsubst{r}{s_1}{s_2} \in \lang{\lreplace{r}{r_1}{r_2}}$.
+\end{prf}
+
+\begin{prop}[Replacement Correspondence] \label{thm:substcorrespondence}
+  If $s_1 \in \lang{r_1}$ and $s_2 \in \lang{r_2}$ then\\ $\lsubst{r}{s_1}{s_2} \in \lang{\lreplace{r}{r_1}{r_2}}$.
 \end{prop}
-\begin{proof}
-The proposition follows from the definitions of $\textsf{subst}$ and $\textsf{lreplace}$; note that language substitutions over-approximate string substitutions.
-\end{proof}
 
@@
 \end{lem}
-\begin{proof}
-  The only non-trivial cases are S-T-Replace and S-T-Concat. The first follows from Proposition \ref{thm:total}, and the other
-  follows from the fact that lhead and ltail always produce well-formed expressions.
-\end{proof}
 
@@
 \end{thm}
-\begin{proof}
+\begin{prf}
 By induction on the typing relation.
@@
 and the S-T-Replace case appeals to Proposition \ref{thm:substcorrespondence}.
-\end{proof}
+\end{prf}
 
@@
 
-\begin{lem}[Canonical Forms for String Fragment of $\lambdas$]\label{thm:cfs}
+\begin{lem}[Normalization for String Fragment of $\lambdas$]\label{thm:cfs}
   If $\emptyset \vdash e : \stringin{r}$ and $e \sreduces v$ then $v=\rstr{s}$. 
 \end{lem}
-\begin{proof}
-The proof is again by induction on the typing relation, following straightforwardly.
-\end{proof}
 
@@
 \end{thm}
-\begin{proof}
+\begin{prf}
   The theorem follows directly from type safety, canonical forms for $\lambdas$, and inversion of \textsc{S-T-Stringin-I}, the typing rule for terms of the form $\rstr{s}$.
-\end{proof}
+\end{prf}
 
@@
 <$\tau$> ::= $\tau \rightarrow \tau$ | $\str$ | $\regex$ \hfill target types
-
-<$\iota$> ::= $x$ | $\lambda x . \iota$ | $\iota(\iota)$ \hfill target terms \alt
-$\tstr{s}$ | $\tconcat{\iota}{\iota}$ | $\pstrcase{\iota}{\iota}{x,y.\iota}$ \alt
-  $\rx{r}$ | $\tcheck{\iota}{\iota}{\iota}{\iota}$ | $\preplace{\iota}{\iota}{\iota}$
-
-  <$\dot{v}$> ::= $\lambda x . \iota$ | $\tstr{s}$ | $\rx{r}$ \hfill target values
-
+\vspace{-8px}
+
+<$\dot{v}$> ::= $\lambda x . \iota$ | $\tstr{s}$ | $\rx{r}$ \hfill target values ($s \in \Sigma^{*}$)
+\vspace{-8px}
+
+<$\iota$> ::= $\dot{v}$ | $x$ | $\iota(\iota)$ \hfill target terms \alt
+   $\tconcat{\iota}{\iota}$ | $\pstrcase{\iota}{\iota}{x,y.\iota}$ \alt
+   $\tcheck{\iota}{\iota}{\iota}{\iota}$ | $\preplace{\iota}{\iota}{\iota}$
 \end{grammar}
-\caption{Syntax for the target language, $\lambdap$, containing strings and statically constructed regular expressions.}
+\vspace{-5px}
+\caption{Syntax of the target language, $\lambdap$, containing strings and statically constructed regular expressions.}
 \label{fig:lcsSyntax}
@@
 is preserved under translation into $\lambdap$, which we now define.
+
+\begin{figure}[t]\label{fig:lambdap}
+\scriptsize
+$\fbox{\inferrule{}{\tctx \vdash \iota : \tau}}$
+~~~~$\tctx ::= \emptyset \pipe \tctx, x : \tau$
+\begin{mathpar} 
+\inferrule[P-T-Var]
+{ x:\tau \in \tctx }
+{ \tctx \vdash x:\tau }
+
+\inferrule[P-T-Abs]
+{\tctx, x : \tau_1 \vdash \iota_2 : \tau_2}
+{\tctx \vdash \lambda x.\iota_2 : \tau_1 \rightarrow \tau_2}
+
+\inferrule[P-T-App]
+{\tctx \vdash \iota_1 : \tau_2 \rightarrow \tau \\ \tctx \vdash \iota_2 : \tau_2}
+{\tctx \vdash \iota_1(\iota_2) : \iota}
+
+\inferrule[P-T-String]
+{ \ }
+{\tctx \vdash \tstr{s} : \str}
+
+\inferrule[P-T-Regex]
+{ \ }
+{\tctx \vdash \rx{r} : \regex}
+
+\inferrule[P-T-Concat]
+{\tctx \vdash \iota_1 : \str \\ \tctx \vdash \iota_2 : \str}
+{\tctx \vdash \tconcat{\iota_1}{\iota_2} : \str}
+
+\inferrule[P-T-Case]
+{\tctx \vdash \iota_1 : \str \\ \tctx \vdash \iota_2 : \tau \\ \tctx, x:\str, y:\str \vdash \iota_3:\tau}
+{\tctx \vdash \pstrcase{\iota_1}{\iota_2}{x,y.\iota_3} : \tau}
+
+\inferrule[P-T-Replace]
+{\tctx \vdash \iota_1 : \regex \\ \tctx \vdash \iota_2 : \str \\ \tctx \vdash \iota_3 : \str }
+{\tctx \vdash \preplace{\iota_1}{\iota_2}{\iota_3} : \str}
+
+\inferrule[P-T-Check]
+{\tctx \vdash \iota_r : \regex \\ \tctx \vdash \iota_1 : \str \\ \tctx \vdash \iota_2 : \sigma \\ \tctx \vdash \iota_3 : \sigma}
+{\tctx \vdash \tcheck{\iota_r}{\iota_1}{\iota_2}{\iota_3} : \sigma}
+\end{mathpar}
+\vspace{-10pt}
+\caption{Typing rules for $\lambdap$.
+The typing context $\tctx$ is standard.}
+\label{fig:slambdap}
+\end{figure}
 
@@
 \renewcommand{\grammarlabel}[2]{#1\hfill#2}
-
-\begin{figure}[t]\label{fig:lambdap}
-\small
-$\fbox{\inferrule{}{\tctx \vdash \iota : \tau}}$
-~~~~$\tctx ::= \emptyset \pipe \tctx, x : \tau$
-
-\begin{mathpar}
-\inferrule[P-T-Var]
-{ x:\tau \in \tctx }
-{ \tctx \vdash x:\tau }
-
-\inferrule[P-T-Abs]
-{\tctx, x : \tau_1 \vdash \iota_2 : \tau_2}
-{\tctx \vdash \lambda x.\iota_2 : \tau_1 \rightarrow \tau_2}
-
-\inferrule[P-T-App]
-{\tctx \vdash \iota_1 : \tau_2 \rightarrow \tau \\ \tctx \vdash \iota_2 : \tau_2}
-{\tctx \vdash \iota_1(\iota_2) : \iota}
-
-\inferrule[P-T-String]
-{ \ }
-{\tctx \vdash \tstr{s} : \str}
-
-\inferrule[P-T-Regex]
-{ \ }
-{\tctx \vdash \rx{r} : \regex}
-
-\inferrule[P-T-Concat]
-{\tctx \vdash \iota_1 : \str \\ \tctx \vdash \iota_2 : \str}
-{\tctx \vdash \tconcat{\iota_1}{\iota_2} : \str}
-
-\inferrule[P-T-Case]
-{\tctx \vdash \iota_1 : \str \\ \tctx \vdash \iota_2 : \tau \\ \tctx, x:\str, y:\str \vdash \iota_3:\tau}
-{\tctx \vdash \pstrcase{\iota_1}{\iota_2}{x,y.\iota_3} : \tau}
-
-\inferrule[P-T-Replace]
-{\tctx \vdash \iota_1 : \regex \\ \tctx \vdash \iota_2 : \str \\ \tctx \vdash \iota_3 : \str }
-{\tctx \vdash \preplace{\iota_1}{\iota_2}{\iota_3} : \str}
-
-\inferrule[P-T-Check]
-{\tctx \vdash \iota_r : \regex \\ \tctx \vdash \iota_1 : \str \\ \tctx \vdash \iota_2 : \sigma \\ \tctx \vdash \iota_3 : \sigma}
-{\tctx \vdash \tcheck{\iota_r}{\iota_1}{\iota_2}{\iota_3} : \sigma}
-\end{mathpar}
-\caption{Typing rules for $\lambdap$.
-The typing context $\tctx$ is standard.}
-\label{fig:slambdap}
-\end{figure}
-
-\begin{figure}[t]
-\small
-$\fbox{\inferrule{}{\iota \treduces \dot{v}}}$
-
-\begin{mathpar}
-\inferrule[P-E-Abs]
-{ \ }
-{\lambda x.e \sreduces \lambda x.e}
-
-\inferrule[P-E-App]
-{ \iota_1 \sreduces \lambda x . \iota_3 \\  \iota_2 \sreduces \dot{v}_2 \\ [\dot{v}_2 / x] \iota_3 \sreduces \dot{v}_3}
-{ \iota_1(\iota_2) \sreduces \dot{v}_3}
-
-\inferrule[P-E-Str]
-{ \ }
-{\tstr{s} \treduces \tstr{s}}
-
-\inferrule[P-E-Rx]
-{ \ }
-{\rx{r} \treduces \rx{r}}
-
-\inferrule[P-E-Concat]
-{\iota_1 \treduces \tstr{s_1} \\ \iota_2 \treduces \tstr{s_2}} 
-{\tconcat{\iota_1}{\iota_2} \treduces \tstr{s_1 s_2}} % ???
-
-\inferrule[P-E-Case-$\epsilon$]
-{
-  \iota_1 \treduces \tstr{ \epsilon } \\
-  \iota_2 \treduces \dot{v_2} \\
-}
-{
-  \pstrcase{\iota_1}{\iota_2}{x,y.\iota_3} \treduces \dot{v_2}
-}
-
-\inferrule[P-E-Case-Concat]
-{
-  \iota_1 \treduces \tstr{as} \\
-  [\tstr{a},\tstr{s}/x,y] \iota_3 \treduces \dot{v} 
-}
-{
-  \pstrcase{\iota_1}{\iota_2}{x,y.\iota_3} \treduces \dot{v}
-}
-
-\inferrule[P-E-Replace]
-{\iota_1 \treduces \rx{r} \\ \iota_2 \treduces \tstr{s_2} \\ \iota_3 \treduces \tstr{s_3} \\ \lsubst{r}{s_2}{s_3} = s} 
-{\preplace{\iota_1}{\iota_2}{\iota_3} \treduces \tstr{s}}
-
-\inferrule[P-E-Check-OK]
-{
-  \iota_r \treduces \rx{r} \\
-  \iota \treduces \tstr{s} \\ 
-s \in \lang{r} \\ 
-\iota_1 \treduces \dot{v_1} 
-}
-{
-  \tcheck{\iota_r}{\iota}{\iota_1}{\iota_2} \treduces \dot{v_1}
-}
-
-\inferrule[P-E-Check-NotOK]
-{\iota_r \treduces \rx{r} \\ \iota \treduces \tstr{s} \\ s \not \in \lang{r} \\ \iota_2 \treduces \dot{v_2}}
-{\tcheck{\iota_r}{\iota}{\iota_1}{\iota_2} \treduces \dot{v_2}}
-\end{mathpar}
-\caption{Big step semantics for $\lambdap$}.
-\label{fig:dlambdap}
-\end{figure}
 
@@
 
+\begin{thm}[Translation Correctness]\label{thm:trcorrect}
+  If $\tctx \vdash e : \sigma$ then 
+  there exists an $\iota$ such that $\trden{e} = \iota$
+  and $\trden{\tctx} \vdash \iota : \trden{\sigma}$.
+  Furthermore, $e \sreduces v$ and
+  $\iota \treduces \dot{v}$ such that
+  $\trden{v} = \dot{v}$.
+\end{thm}
+\begin{prf}
+The proof proceeds by induction on the typing relation for $e$.
+We choose an $\iota$ based on the syntactic form in $\lambdap$ corresponding to the form under consideration (e.g.
+we  choose {\sf replace} when considering {\sf sreplace}).
+The proof proceeds by our type safety theorems and an appeal to the induction hypothesis.
+\end{prf}
+
+\begin{figure}[t]
+\scriptsize
+$\fbox{\inferrule{}{\iota \treduces \dot{v}}}$
+
+\begin{mathpar}
+\inferrule[P-E-Abs]
+{ \ }
+{\lambda x.\iota \sreduces \lambda x.\iota}
+
+\inferrule[P-E-App]
+{ \iota_1 \sreduces \lambda x . \iota_3 \\  \iota_2 \sreduces \dot{v}_2 \\ [\dot{v}_2 / x] \iota_3 \sreduces \dot{v}_3}
+{ \iota_1(\iota_2) \sreduces \dot{v}_3}
+
+\inferrule[P-E-Str]
+{ \ }
+{\tstr{s} \treduces \tstr{s}}
+
+\inferrule[P-E-Rx]
+{ \ }
+{\rx{r} \treduces \rx{r}}
+
+\inferrule[P-E-Concat]
+{\iota_1 \treduces \tstr{s_1} \\ \iota_2 \treduces \tstr{s_2}} 
+{\tconcat{\iota_1}{\iota_2} \treduces \tstr{s_1 s_2}} % ???
+
+\inferrule[P-E-Case-$\epsilon$]
+{
+  \iota_1 \treduces \tstr{ \epsilon } \\
+  \iota_2 \treduces \dot{v_2} \\
+}
+{
+  \pstrcase{\iota_1}{\iota_2}{x,y.\iota_3} \treduces \dot{v_2}
+}
+
+\inferrule[P-E-Case-Concat]
+{
+  \iota_1 \treduces \tstr{as} \\
+  [\tstr{a},\tstr{s}/x,y] \iota_3 \treduces \dot{v} 
+}
+{
+  \pstrcase{\iota_1}{\iota_2}{x,y.\iota_3} \treduces \dot{v}
+}
+
+\inferrule[P-E-Replace]
+{\iota_1 \treduces \rx{r} \\ \iota_2 \treduces \tstr{s_2} \\ \iota_3 \treduces \tstr{s_3} \\ \lsubst{r}{s_2}{s_3} = s} 
+{\preplace{\iota_1}{\iota_2}{\iota_3} \treduces \tstr{s}}
+
+\inferrule[P-E-Check-OK]
+{
+  \iota_r \treduces \rx{r} \\
+  \iota \treduces \tstr{s} \\ 
+s \in \lang{r} \\ 
+\iota_1 \treduces \dot{v_1} 
+}
+{
+  \tcheck{\iota_r}{\iota}{\iota_1}{\iota_2} \treduces \dot{v_1}
+}
+
+\inferrule[P-E-Check-NotOK]
+{\iota_r \treduces \rx{r} \\ \iota \treduces \tstr{s} \\ s \not \in \lang{r} \\ \iota_2 \treduces \dot{v_2}}
+{\tcheck{\iota_r}{\iota}{\iota_1}{\iota_2} \treduces \dot{v_2}}
+\end{mathpar}
+\vspace{-10px}
+\caption{Big step semantics for $\lambdap$}.
+\label{fig:dlambdap}
+\end{figure}
+
+\subsubsection{Preservation of Security}
+
+Finally, our main result establishes that correctness of $\lambdas$ is preserved under the translation into $\lambdap$. \renewcommand*{\proofname}{Proof}
+
+\begin{thm}[Correctness of Input Sanitation for Translated Terms]\label{thm:main}
+  If $\trden{e} = \iota$ and $\emptyset \vdash e : \stringin{r}$ then $\iota \sreduces \tstr{s}$
+  for $s \in \lang{r}$.
+\end{thm}
+\begin{prf}
+  By Theorem \ref{thm:trcorrect} and the rules given, $\iota \sreduces \tstr{s}$ implies that $e \sreduces \strin{s}$.
+  Theorem \ref{thm:scorrect} together with the assumption that $e$ is well-typed implies that $s \in \lang{r}$.
+\end{prf}
+
+%\begin{figure}
+%\begin{lstlisting}
+%output of successful compilation
+%\end{lstlisting}
+%\caption{Output of successful compilation.}
+%\end{figure}
+%
+%\begin{figure}
+%\begin{lstlisting}
+%output of failed compilation
+%\end{lstlisting}
+%\caption{Output of failed compilation.}
+%\end{figure}
+%
+
+%\subsection{An Example}\label{whatev}
+
 \begin{figure}[ht]
-\small
+\scriptsize
 $\fbox{\inferrule{}{\trden{\sigma}=\tau}}$
-\begin{mathpar}
+\vspace{-25px}\begin{mathpar}\hspace{20px}
 \inferrule[Tr-T-String]
@@
 $\fbox{\inferrule{}{\trden{\Psi}=\Theta}}$
-\begin{mathpar}
+\vspace{-25px}\begin{mathpar}\hspace{20px}
 \inferrule[Tr-T-Context-Emp]
@@
 
-\begin{thm}[Translation Correctness]\label{thm:trcorrect}
-  If $\tctx \vdash e : \sigma$ then 
-  there exists an $\iota$ such that $\trden{e} = \iota$
-  and $\trden{\tctx} \vdash \iota : \trden{\sigma}$.
-  Furthermore, $e \sreduces v$ and
-  $\iota \treduces \dot{v}$ such that
-  $\trden{v} = \dot{v}$.
-\end{thm}
-\begin{proof}
-The proof proceeds by induction on the typing relation for $e$.
-We choose an $\iota$ based on the syntactic form in $\lambdap$ corresponding to the form under consideration (e.g.
-we  choose {\sf replace} when considering {\sf sreplace}).
-The proof proceeds by our type safety theorems and an appeal to the induction hypothesis.
-\end{proof}
-
-\subsubsection{Preservation of Security}
-
-Finally, our main result establishes that correctness of $\lambdas$ is preserved under the translation into $\lambdap$. \renewcommand*{\proofname}{Proof}
-
-\begin{thm}[Correctness of Input Sanitation for Translated Terms]\label{thm:main}
-  If $\trden{e} = \iota$ and $\emptyset \vdash e : \stringin{r}$ then $\iota \sreduces \tstr{s}$
-  for $s \in \lang{r}$.
-\end{thm}
-\begin{proof}
-  By Theorem \ref{thm:trcorrect} and the rules given, $\iota \sreduces \tstr{s}$ implies that $e \sreduces \strin{s}$.
-  Theorem \ref{thm:scorrect} together with the assumption that $e$ is well-typed implies that $s \in \lang{r}$.
-\end{proof}
-
-%\begin{figure}
-%\begin{lstlisting}
-%output of successful compilation
-%\end{lstlisting}
-%\caption{Output of successful compilation.}
-%\end{figure}
-%
-%\begin{figure}
-%\begin{lstlisting}
-%output of failed compilation
-%\end{lstlisting}
-%\caption{Output of failed compilation.}
-%\end{figure}
-%
-
-%\subsection{An Example}\label{whatev}
-
-\section{Implementation in {Atlang}}\label{atlang}
+\section{Implementation in \tt{atlang}}\label{atlang}
 In the previous section, we specified a type system and a translation semantics to a language containing only strings and regular expressions.
@@
 
-\newcommand{\lstinlinep}[1]{\lstinline[language=Python,basicstyle=\ttfamily\small]{#1}}
+\newcommand{\lstinlinep}[1]{\lstinline[language=Python,basicstyle=\ttfamily\scriptsize]{#1}}
 
@@
 
-\begin{figure}[ht]\small\begin{lstlisting}
+\begin{figure}[t]\scriptsize\begin{lstlisting}
 from atlib import fn, stringin
@@
 \end{lstlisting}
-\vspace{-10px}
-\caption{Regular string types in atlang, a library that enables static type checking for Python. }\label{fig:atexample}
+%\vspace{-10px}
+\caption{Regular string types in {\tt atlang}}\label{fig:atexample}
 \end{figure}
@@
 \end{lstlisting}
-\vspace{-10px}
+%\vspace{-10px}
 \caption{The output of compilation of Figure \ref{fig:atexample} (at the terminal, \texttt{atc figure9.py}, or just-in-time).}
@@
 
-  def syn_Method_replace(self, ctx, node):
-    [rx, exp] = node.args
-    if not isinstance(rx, ast.Str):
-      raise atlang.TypeError("...", node)
-    rx = rx.s
-    exp_t = ctx.syn(exp)
-    if not isinstance(exp_t, stringin):
-      raise atlang.TypeError("...", node)
-    exp_rx = exp_t.idx
-    return stringin[lreplace(self.idx, rx, exp_rx)]
-
-  def trans_Method_replace(self, ctx, node):
-    return astx.quote(
-      """__import__(re); re.sub(%0, %1, %2)""",
-      astx.Str(s=node.args[0]),
-      astx.copy(node.func.value),
-      astx.copy(node.args[1]))
-
-  # check and strcase omitted
+  # replace, check and strcase omitted. See TR.
 
@@
 \end{lstlisting}
-\vspace{-10px}
-\caption{Implementation of the \texttt{stringin} type constructor in atlang.}
+%\vspace{-10px}
+\caption{Implementation of the \texttt{stringin} type constructor in {\tt atlang}.}
 \label{fig:impl}
@@
 This work was supported by the National Security Agency
-lablet contract \#H98230-14-C-0140.
+lablet contract \#H98230-14-C-0140 and the National Science Foundation under NSF CNS-1035800.
+%The views and conclusions contained in this document are those of the author and should not be inter- preted as representing the official policies, either expressed or implied, of any sponsoring institution or government.
+Any opinions, findings, and conclusions or recommendations expressed in this publication are those of the author(s) and do not necessarily reflect the views of any sponsoring institution or government.
+
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
 \bibliographystyle{abbrv}
-
+\renewcommand*{\bibfont}{\footnotesize}
+\setlength{\bibsep}{0pt plus .25pt minus .25pt}
+{
+\vspace*{-1.5mm}
 \bibliography{research}
-
+}
 % The bibliography should be embedded for final submission.

### 2014-12-02 att-pldi15/att-pldi15.tex

@@
 In contrast, in @$\lambda$, fragment providers need not provide the semantics with mechanized specfications or proofs to benefit from rigorous modular reasoning principles.
-Instead, under a fixed tycon context, the calculus can be reasoned about like a very small type-directed compiler \cite{tarditi+:til-OLD,conf/pldi/Chlipala07}.
+Instead, under a fixed tycon context, the calculus can be reasoned about like a very small type-directed compiler \cite{tarditi+:til-OLD,conf/pldi/Chlipala07,Dave:2003:CVB:966221.966235}.
 Errors in reasoning can only lead to failure at typechecking time, via our chief contribution: a novel form of \emph{translation validation} \cite{Pnueli-Siegel-Singerman98}.
@@
 To eliminate even these localized failure modes, we plan to introduce \emph{optional} specification and proof mechanization into the SL (by basing it on a dependently typed language like Coq, rather than ML).
-Because types are values in the SL, the SL can be seen as building on the concept of \emph{type-level computation}.
+Because types are values in the SL, the SL can be seen as building on the concept of \emph{type-level computation}\todo{find appropriate citation}.
 %We also hope to mechanize the metatheory of @$\lambda$ itself in future work.%Every combination of fragments is a new dialect, and must be reasoned about monolithically.
@@
 In contrast, modules and ADTs can only specify a fixed number of operations, and each must have function type.
-Note also that these are not competing mechanisms: we did not specify quantification over external types here for simplicity, but we conjecture that this is entirely complementary and thus @$\lambda$ could serve as the core of a language with an ML-style module system and parametric polymorphism.
+Note also that these are not competing mechanisms: we did not specify quantification over external types here for simplicity, but we conjecture, based on the finding that polymorphism is conservative over simple types, \cite{breazu1990polymorphism}, that this is  complementary and thus @$\lambda$ could serve as the core of a language with an ML-style module system and polymorphism.
 Another related direction is to explore \emph{tycon functors}, which would abstract over tycons with the same signature to support modularly tunable cost semantics.
@@
 \section{Conclusion}
-\noindent We have specified a remarkably simple but surprisingly powerful typed translation semantics, @$\lambda$, where new type constructors can be introduced ``from within''.
+\noindent We have specified a simple but surprisingly powerful typed translation semantics, @$\lambda$, where new type constructors can be introduced ``from within''.
 The corresponding operator constructors determine types and translations ``actively'', i.e.
@@
 A simple form of translation validation based on existing notions of type abstraction  ensures that opcons associated with one tycon maintain translation independence from all others, guaranteeing that a wide class of important properties can be reasoned about modularly.
-Implementation is ongoing.
+Implementation in several settings is ongoing.
 

### 2014-12-02 sciunit-icse14/sciunit-icse14.tex

@@
 \titleformat{\section}
- {\secfnt}{\thesection}{1em}{\secfnt\MakeUppercase}
+ {\secfnt}{\thesection.}{1em}{\secfnt\MakeUppercase}
  \vspace{10pt}

### 2014-12-02 sanitation-psp14/excluded_fig11.py
Added content excluded from paper, to be included in the TR.

@@ -1,0 +1,23 @@
+  def syn_Method_replace(self, ctx, node):                                      
+    [rx, exp] = node.args                                                       
+    if not isinstance(rx, ast.Str):                                             
+      raise atlang.TypeError("...", node)                                       
+    rx = rx.s                                                                   
+    exp_t = ctx.syn(exp)                                                        
+    if not isinstance(exp_t, stringin):                                         
+      raise atlang.TypeError("...", node)                                       
+    exp_rx = exp_t.idx                                                          
+    return stringin[lreplace(self.idx, rx, exp_rx)]                             
+                                                                                
+  def trans_Method_replace(self, ctx, node):                                    
+    return astx.quote(                                                          
+      """__import__(re); re.sub(%0, %1, %2)""",                                 
+      astx.Str(s=node.args[0]),                                                 
+      astx.copy(node.func.value),                                               
+      astx.copy(node.args[1]))                                                  
+
+
+
+
+also this text:
+The regular expressions describing the first character and remaining characters of a matching string in the language $\lang{r}$ must be determined to correctly specify the static semantics. For instance, the first character of a string in the language $\lang{(A\cdot B\cdot C)+(D\cdot E\cdot F)}$ is in the language $\lang{A+D}$ and the remaining characters are in the language $\lang{(B\cdot C)+(E \cdot F)}$. 

### 2014-12-02 sanitation-psp14/final.tex
Final PSP14 main paper

@@
 %\usepackage{enumitem}
-\usepackage{enumerate}
+\usepackage{enumerate}%
 
@@
 Web applications must ultimately command systems like web browsers and database engines using strings.
-Strings derived from improperly sanitized user input  can thus be a vector for command injection attacks.
-In this paper, we introduce \emph{regular string types}, which classify strings known statically to be in a specified regular language.
-These types come equipped with common operations like concatenation, substitution and coercion, so they  can be used to implement, in a conventional manner, the portions of a web application or application framework that must directly construct command strings.
-Simple type annotations at key interfaces can be used to statically verify that sanitization has been performed correctly without introducing redundant run-time checks.
-We specify this type system in a minimal typed lambda calculus, $\lambda_{RS}$.
+Strings derived from improperly sanitized user input  can as a result be a vector for command injection attacks.
+In this paper, we introduce \emph{regular string types}, which classify strings constrained statically to be in a regular language specified by a regular expression.
+Regular strings support standard string operations like concatenation and substitution, as well as safe coercions, so they  can be used to implement, in an essentially conventional manner, the pieces of a web application or framework that  handle strings arising from user input.
+Simple type annotations at function interfaces can be used to statically verify that sanitization has been performed correctly without introducing redundant run-time checks.
+We specify this type system first as a minimal typed lambda calculus, $\lambda_{RS}$.
 
@@
 We support this with two contributions.
-First, we specify a translation from $\lambda_{RS}$ to a language fragment containing only standard strings and regular expressions.
-Second, taking Python as a language with these constructs, we implement the type system together with the translation as a library using \texttt{atlang}, an extensible static type system for Python being developed by the authors.
+First, we specify a translation from $\lambda_{RS}$ to a calculus with only standard strings and regular expressions.
+Then, taking Python as a language with these constructs, we implement the type system together with the translation as a library using \texttt{atlang}, an extensible static type system for Python.
 \end{abstract}
@@
 For example, web browsers are controlled using HTML and Javascript sent from a server as a string, and database engines execute SQL queries also sent as strings.
-When these commands include data derived from user input, care must be taken to ensure that the user cannot  subvert the intended command by carefully crafting the data they send.
+When these commands include substrings derived from user input, care must be taken to ensure that the user cannot  subvert the intended command.
 For example, a  SQL query constructed using string concatenation exposes a SQL injection vulnerability:
@@
 'SELECT * FROM users WHERE name="' + name + '"'
-\end{lstlisting}
+\end{lstlisting}\vspace{-3px}
 If a malicious user enters the name \lstinline{'"; DROP TABLE users --'}, the entire database could be erased.
 
-To avoid this problem, the program must \emph{sanitize} user input.
-For example, in this case, the developer (or, more often, a framework) might define a function \verb|sanitize| that escapes double quotes and existing backslashes   with a backslash, which SQL treats safely.
+To avoid this problem, developers are cautioned to \emph{sanitize} user input.
+For example, in this case, the developer (or, more often, a framework) might define a function \verb|sanitize| that escapes double quotes and existing backslashes   with a backslash, which SQL treats securely.
 Alternatively, it might HTML-encode special characters, which would head off both SQL injection attacks and cross-site scripting attacks.
-Guaranteeing that user input has already been sanitized  before it is used to construct a command is challenging, especially because  sanitization and command construction may occur in separately developed and maintained components.
+\emph{Guaranteeing} that user input has already been sanitized  before it is used to construct a command is challenging, especially because  sanitization and command construction may occur in separately developed and maintained components.
 %is often performed in a different component of a program than where commands are ultimately constructed.% Note, for example, that \verb|sanitize| is not idempotent, so it should only be called once.
 
-We observe that many such sanitization techniques can be understood using \emph{regular languages} \cite{cinderella}.
-For example, \verb|name| must be a string in the language described by the regular expression \verb!([^"\]|(\")|(\\))*!
+We observe that many such sanitization techniques can be specified using \emph{regular languages} \cite{cinderella}.
+For example, for the command constructed above to be secure, \verb|name| must be a string in the language described by the regular expression \verb!([^"\\]|(\")|(\\))*!
 -- a sequence of characters other than quotation marks and backslashes; these can only appear escaped.
-This syntax for regular expression patterns can be understood to desugar, in a standard way, to the syntax for regular expressions shown in Figure \ref{fig:regex}, where $r \cdot r$ is sequencing and $r + r$ is  disjunction.
-We will work with this ``core'' for simplicity.
-
-In this paper, we present a static type system that tracks the regular language a string belongs to.
-For example, the output of \verb|sanitize| will be a string in the regular language described by the regular expression above (we identify regular languages by the notation $\mathcal{L}\{r\}$).
-By leveraging closure and decidability properties of regular languages, the type system tracks the language of a string through uses of a number of operations, including \emph{replacement} of substrings matching a given pattern.
-This makes it possible to implement sanitation functions -- like the one just described -- in a conventional manner.
-The result is a system where the fact that a string has been {correctly} sanitized is manifest in its type.
-Missing calls to sanitization functions are detected statically, and, importantly, so are \emph{incorrectly implemented sanitization functions} (i.e.
+This conventional syntax for regular expression patterns can be understood to desugar, in a standard way, to the syntax for regular expressions shown in Figure \ref{fig:regex}, where $r \cdot r$ is sequencing and $r + r$ is  disjunction.
+We will work with this ``core'' for simplicity in the remainder.
+
+In this paper, we present a static type system that tracks the regular language a string belongs to (we identify regular languages by the notation $\mathcal{L}\{r\}$).
+For example, the output of \verb|sanitize| will not simply be \verb|string|, but rather \verb|stringin[r]|, where \verb|r| is the regular expression above.
+By leveraging closure and decidability properties of regular languages, the type system will be able to closely track the language a string belongs to through uses of a number of operations, including \emph{replacement} of substrings matching a given pattern.
+This makes it possible to implement sanitization functions -- like the one just described -- in a conventional manner.
+The result is a system where the fact that a string has been {correctly} sanitized becomes manifest in its type.
+Missing calls to sanitization functions can thus be detected statically, and, crucially, so can \emph{incorrectly implemented sanitization functions} (i.e.
 these functions need not be trusted).
-These guarantees require run-time checks only when going from less precise to more precise types (e.g.
-along the interfaces with external systems or components that are not under the control of this type system).
+These guarantees require run-time checks only when going from less precise to more precise types.
 %implementing and statically checking input sanitation techniques.
@@
 Waiting for a language designer to build this feature in is unsatisfying in practice.
-Moreover, we also face a ``chicken-and-egg problem'': justifying its inclusion into a commonly used language benefits from empirical case studies, but these are  difficult to conduct if developers have no way to use the abstraction in real-world code.
+%Moreover, we also face a ``chicken-and-egg problem'': justifying its inclusion into a commonly used language benefits from empirical case studies, but these are  difficult to conduct if developers have no way to use the abstraction in real-world code.
 We take the position that a better path forward for the community is to work within a programming language where such type system fragments can be introduced modularly and orthogonally, as libraries.
@@
 In Sec.
-\ref{atlang}, we show how to implement the type system fragment we have specified  using \texttt{atlang}, an extensible static type system implemented as a library inside Python.
+\ref{atlang}, we show how to implement the type system and translation from Sec.
+\ref{calculus} using \texttt{atlang}, an extensible static type system implemented as a library inside Python.
 \texttt{atlang} leverages local type inference to control the semantics of literal forms, so regular string types can be introduced using string literals without any run-time overhead.
@@
 \begin{itemize}
-\item Section \ref{sec:rs} describes $\lambda_{RS}$. We also give proof outlines of  type safety and correctness theorems and relevant propositions about regular languages.
-\item Section \ref{sec:p} describes a simple target language, $\lambdap$, with a minimal regular expression library. In Section \ref{atlang}, we will take Python to be such a language.
-\item Section \ref{sec:tr} describes the translation from $\lambdas$ to $\lambdap$ and ensures the correctness result for \ref{sec:rs} is preserved under this translation. 
+\item Sec. \ref{sec:rs} describes $\lambda_{RS}$ and its metatheory.% We also give proof outlines of  type safety and correctness theorems and relevant propositions about regular languages.
+\item Sec. \ref{sec:p} describes a simple target language, $\lambdap$, with a minimal regular expression library. In Section \ref{atlang}, we will take Python to be such a language.
+\item Sec. \ref{sec:tr} describes the translation from $\lambdas$ to $\lambdap$ and ensures the correctness result from Sec. \ref{sec:rs} is preserved under this translation. 
 \end{itemize}
 
-The accompanying technical report \cite{psptr} contains more detailed proofs, further discussion of string substitution and language replacement, and
-a single sheet containing Figures \ref{fig:glambdas}-\ref{fig:tr} for the reader's convenience.
+The accompanying technical report \cite{psptr} contains more detailed proofs and further discussion of  design choices.% Figures \ref{fig:glambdas}-\ref{fig:tr} for the reader's convenience.
 
 \subsection{The Language of Regular Strings}\label{sec:rs}
-In this section, we define a minimal typed lambda calculus with regular string types called $\lambda_{RS}$.
-The syntax of $\lambda_{RS}$ is specified in Figure \ref{fig:glambdas}, its static semantics in Figure \ref{fig:slambdas} and its evaluation semantics in Figure \ref{fig:dlambdas}.
-
-There are two type constructors in $\lambdas$, $\rightarrow$ and $\textsf{stringin}$.
+In this section, we define a typed lambda calculus with regular string types called $\lambda_{RS}$.
+Its syntax is specified in Figure \ref{fig:glambdas}, its static semantics in Figure \ref{fig:slambdas} and its evaluation semantics, given here in big-step style, in Figure \ref{fig:dlambdas}.
+
+There are two type constructors in $\lambdas$: $\rightarrow$ and $\textsf{stringin}$.
 Arrow types classify functions, which are introduced via lambda abstraction, $\lambda x.e$, and can be applied, written $e(e)$, in the usual way \cite{pfpl}.
@@
 Values of such regular string types take the form $\strin{s}$, where $s$ is a string (i.e.
-$s \in \Sigma^\star$, defined in the usual way).
+$s \in \Sigma^\star$, where the Kleene star is defined in the usual way).
 The rule \textsc{S-T-Stringin-I} requires that $s \in \lang{r}$.
@@
 $\lambdas$ provides several familiar operations on strings. The type system  relates these operations over strings to corresponding operations over the  regular languages they belong to. 
-Since these operations over regular languages are known to be closed and decidable, we can use these operations as a basis for static analysis of sanitation protocols (we will see a complete example in Section \ref{atlang}).
+Since these operations over regular languages are known to be closed and decidable, we can use these operations as a basis for implementing and reasoning about sanitation protocols, as we will discuss below.
 
@@
 
-\subsubsection{Concatenation and String Decomposition}
+\subsubsection{Concatenation}
 
@@
 
+\subsubsection{String Decomposition}
 Whereas concatenation allows the construction of large strings from smaller strings,  $\strcase{e}{e_0}{x,y.e_1}$  allows the decomposition, or elimination, of large strings into smaller strings.
@@
 
+The typing rule \textsc{S-T-Case} must determine a suitable type for the head and tail of the string.
 The regular expression recognizing any one-character prefix of the strings in $\lang{r}$ is easily defined.
@@
 \begin{defn}[Definition of $\lhead{r}$] 
-$\lhead{r}$ is defined in terms of the structure of $r$:
 {\footnotesize
 \begin{align*}
-\lhead{r}                 &= \lhead{r, \epsilon} \\                             
+\lhead{r}                 &= \lhead{r, \epsilon}
+\end{align*}}We use a two-argument auxiliary $\lhead{r, r'}$ in the definition of $\lhead{r}$ because when $r = q^*\cdot r'$,
+$\lhead{r}$ needs to ``remember" $r'$ in case $q$ is iterated zero times:
+{\footnotesize
+\begin{align*}
 \lhead{\epsilon, r'}      &= \epsilon \\                                        
@@
 \end{defn}
-We use a two-argument auxiliary $\lhead{r, r'}$ in the definition of $\lhead{r}$ because when $r = q^*\cdot t$,
-$\lhead{r}$ needs to ``remember" $t$ in case $q$ is iterated zero times.
 
@@
 Language inclusion is decidable.
-As a result, the rule \textsc{S-E-SafeCoerce} does not need to perform any checks.
+As a result, the rule \textsc{S-E-SafeCoerce} does not need to perform any dynamic checks.
 The rule
@@
 This can be seen as a form of subtyping for regular string types \cite{gunter, fulton13, xhaskell2}.
-In practice, subtyping is crucial to the usability of our system due to the nature of the replacement operator.
-For simplicity, we do not include this operator; however, subtyping for regular strings is present in previous treatments of our work \cite{fulton13}.
+In practice, subtyping will be crucial to the usability of our system due to the nature of the replacement operator.
+For simplicity, we do not specify subtyping in our core calculus; however, subtyping for regular strings is present in previous treatments of our work \cite{fulton13}.
 The implementation discussed in Section \ref{atlang} also provides subtyping between regular string types.
@@
 It behaves analogously to \lstinline{str_replace} in PHP and \lstinline{String.replace} in Java, differing in that the replacement pattern $r$ must be statically given.
-The evaluation rule \textsc{S-E-replace} is defined in terms of $\lsubst{r}{s_1}{s_2}$.
-For simplicity, we do not define replacement in this paper; see \cite{psptr}.
-
-\begin{defn}[$\tt{subst}$]
-$\lsubst{r}{s_1}{s_2} = s$ produces a string $s$ in which all substrings of $s_1$ in $\lang{r}$ are replaced with $s_2$.
+The evaluation rule \textsc{S-E-replace} is defined in terms of the metafunction $\lsubst{r}{s_1}{s_2}$.
+
+\begin{defn}[$\sf{replace}$]
+$\lsubst{r}{s_1}{s_2} = s$ such that all substrings of $s_1$ in $\lang{r}$ are replaced with $s_2$.
 \end{defn}
-
-The typing rule \textsc{S-T-Replace} requires computing the regular language of the string resulting from this operation given the languages of $s_1$ and $s_2$, written as $\lreplace{r}{r_1}{r_2}={r'}$.
+The typing rule \textsc{S-T-Replace} thus requires computing the regular language of the string resulting from this operation given knowledge of the languages of $s_1$ and $s_2$, written as $\lreplace{r}{r_1}{r_2}={r'}$.
+
 Given an automata-oriented interpretation of regular languages, it may be helpful to think in terms of replacing sub-automata.
-
-A complete definition of ${\sf lreplace}$ would also give an rewrite system with correctness and termination proofs, which is beyond the scope of this paper.
+A complete definition of ${\sf lreplace}$ would consist of a rewrite system based on this intuition, with correctness and termination proofs, which is beyond the scope of this paper.
 Instead, we provide an abstract definition of the operation and state necessary properties.
 
-\begin{defn}[$\tt{lreplace}$]
+\begin{defn}[$\sf{lreplace}$]
   $\lreplace{r}{r_1}{r_2}$ relates $r, r_1,$ and $r_2$ to
@@
 
-\begin{prop}[Closure] \label{thm:total}
-  If $\lang{r}, \lang{r_1}$ and $\lang{r_2}$ are regular languages, then $\lang{\lreplace{r}{r_1}{r_2}}$ is also a regular language.
+% \begin{prop}[Closure] \label{thm:total}
+%   If $\lang{r}, \lang{r_1}$ and $\lang{r_2}$ are regular languages, then $\lang{\lreplace{r}{r_1}{r_2}}$ is also a regular language.
+% \end{prop}
+% \begin{prf}
+% Algorithms for the inclusion problem may be adopted to identify any sublanguage $x \subseteq r$ of $r_1$.
+% The language $x_{pre}$ can be computed by taking total derivatives until the remaining language
+% equals $x$; $x_{post}$ can be computed in a similar way after reversal.
+% Then $r'$ is $x_{pre}r_2x_{post}$.
+% \end{prf}
+
+\begin{prop}[Replacement Correspondence] \label{thm:substcorrespondence}
+  Given strings $s_1 \in \lang{r_1}$ and $s_2 \in \lang{r_2}$ we have $$\lsubst{r}{s_1}{s_2} \in \lang{\lreplace{r}{r_1}{r_2}}$$
 \end{prop}
-\begin{prf}
-Algorithms for the inclusion problem may be adopted to identify any sublanguage $x \subseteq r$ of $r_1$.
-The language $x_{pre}$ can be computed by taking total derivatives until the remaining language
-equals $x$; $x_{post}$ can be computed in a similar way after reversal.
-Then $r'$ is $x_{pre}r_2x_{post}$.
-\end{prf}
-
-\begin{prop}[Replacement Correspondence] \label{thm:substcorrespondence}
-  If $s_1 \in \lang{r_1}$ and $s_2 \in \lang{r_2}$ then\\ $\lsubst{r}{s_1}{s_2} \in \lang{\lreplace{r}{r_1}{r_2}}$.
-\end{prop}
-
-\subsubsection{Safety}
-
-In this section, we establish type soundness for $\lambdas$.
-The theorem relies upon the definitions and propositions given above and some basic properties of regular languages.
+
+\subsubsection{Metatheory of $\lambdas$}
+
+In this section, we establish some basic metatheoretic properties of $\lambdas$.
+These rely upon the definitions and propositions given above and some basic properties of regular languages.
 
@@
 For all strings $s$ and regular expressions $r$, either $s \in \lang{r}$ or $s \not \in \lang{r}$.
-\item 
-Regular languages are closed under reversal.
+%\item 
+%Regular languages are closed under reversal.
 \end{enumerate}
@@
 
-We also need to establish a well-formedness lemma, which simply expresses the totality of the operations over regular expressions used as premises in the rules in Figure \ref{fig:slambdas}. 
-\begin{lem}
-  If $\sctx \vdash e : \stringin{r}$ then $r$ is a well-formed regular expression.
-\end{lem}
-
-Safety for the string fragment of $\lambdas$ requires validating that the type system's definition is justified
-by our theorems about regular languages.
-
-We avoid the most significant issues inherent to proofs related to big-step semantics by avoiding non-termination in our specification.
-The simply typed lambda calculus terminates, and our conservative extension clearly
-terminates because new binding structures are not introduced and termination (i.e.
+% We also need to establish a well-formedness lemma, which simply expresses the totality of the operations over regular expressions used as premises in the rules in Figure \ref{fig:slambdas}. 
+% \begin{lem}
+%   If $\sctx \vdash e : \stringin{r}$ then $r$ is a well-formed regular expression.
+% \end{lem}
+
+Type preservation for $\lambdas$ requires validating that the statics are consistent with the dynamics.
+We give a full proof of type safety for a variant of the calculus with a small step semantics in \cite{psptr}, but for concision, it is more straightforward to explain the semantics with a big step semantics here.
+
+% We avoid the most significant issues inherent to proofs related to big-step semantics by avoiding non-termination in our specification.
+% The simply typed lambda calculus terminates, and our conservative extension clearly
+% terminates because new binding structures are not introduced and termination (i.e.
 decidability) of ${\sf subst}$ and ${\sf lreplace}$ is known.
@@
 
-\begin{thm}[Type Safety.] \label{thm:typesafety}
+\begin{thm}[Type Preservation.] \label{thm:typesafety}
   If $\emptyset \vdash e : \sigma$ 
-  then $e \sreduces v$ and $\emptyset \vdash v : \sigma$.
+  and $e \sreduces v$ then $\emptyset \vdash v : \sigma$.
 \end{thm}
@@
 
-\begin{lem}[Normalization for String Fragment of $\lambdas$]\label{thm:cfs}
-  If $\emptyset \vdash e : \stringin{r}$ and $e \sreduces v$ then $v=\rstr{s}$. 
+\begin{lem}[Canonical Forms for Regular Strings]\label{thm:cfs}
+  If $\emptyset \vdash v : \stringin{r}$ then $v=\rstr{s}$ and $s \in \lang{r}$. 
 \end{lem}
-
+\begin{prf}
+The only typing rule that applies is \textsc{S-T-Stringin-I}. The conclusion is the first premise.
+\end{prf}
 \subsubsection{The Security Theorem}\label{sec:securitythm}
 
-The chief benefit of $\lambdas$ is its security theorem, which states that any value of
-a regular string type is recognized by the regular language corresponding to the regular expression the type is indexed by.
+The chief benefit of $\lambdas$ is its security theorem, which states that any term of
+a regular string type that evaluates to a value will evaluate to a regular string recognized by the regular language corresponding to the regular expression the type is indexed by.
 This is beneficial because this ensure that membership in a regular language known to be secure becomes manifest in the type of the string, rather than being a property that must be established extralinguistically.
@@
 \begin{prf}
-  The theorem follows directly from type safety, canonical forms for $\lambdas$, and inversion of \textsc{S-T-Stringin-I}, the typing rule for terms of the form $\rstr{s}$.
+  The theorem follows directly from type preservation and canonical forms above.
 \end{prf}
@@
 Our next major technical result, stated in Sec. \ref{sec:tr}, establishes that the security property
-is preserved under translation into $\lambdap$, which we now define.
+is preserved under translation into $\lambdap$. The system $\lambdap$ is another straight-forward extension of a simply typed lambda calculus
+with a $\str$ type and a $\regex$ type, as well as 
+some operations -- such as concatenation and replacement -- found in the standard libraries of many programming languages.
+The operations of $\lambdap $ correspond to ``run-time'' versions of the operations performed statically in $\lambdas$ in a way made precise by the translation rules described in the next section. 
+%Unlike $\lambdas$, $\lambdap$ does not statically track the effects of string operations.
+The language $\lambdap$ is so-called because it is reminscent of popular web programming languages,
+such as \textbf{P}ython or \textbf{P}HP, albeit statically typed. We will discuss an implementation within the former in the next section.
+
+The grammar of $\lambdap$ is defined in Figure~\ref{fig:lcsSyntax}.
+The typing rules \textsc{P-T-} are defined in Figure~\ref{fig:slambdap}
+and a big-step semantics is defined by the rules \textsc{P-E-} in Figure~\ref{fig:dlambdap}.
+The semantics are straightforward given the discussion of the corresponding operations in the previous section.
+
+\renewcommand{\grammarlabel}[2]{#1\hfill#2}
+
+\subsubsection{Safety}
 
@@
 \end{figure}
-
-The system $\lambdap$ is another straight-forward extension of a simply typed lambda calculus
-with a $\str$ type and a $\regex$ type, as well as 
-some operations -- such as concatenation and replacement -- found in the standard libraries of many programming languages.
-The operations of $\lambdap $ correspond to ``run-time'' versions of the operations performed statically in $\lambdas$ in a way made precise by the translation rules described in the next section. 
-%Unlike $\lambdas$, $\lambdap$ does not statically track the effects of string operations.
-The language $\lambdap$ is so-called because it is reminscent of popular web programming languages,
-such as \textbf{P}ython or \textbf{P}HP, albeit statically typed. We will discuss an implementation within the former in the next section.
-
-The grammar of $\lambdap$ is defined in Figure~\ref{fig:lcsSyntax}.
-The typing rules \textsc{P-T-} are defined in Figure~\ref{fig:slambdap}
-and a big-step semantics is defined by the rules \textsc{P-E-} in Figure~\ref{fig:dlambdap}.
-The semantics are straightforward given the discussion of the corresponding operations in the previous section.
-
-\renewcommand{\grammarlabel}[2]{#1\hfill#2}
-
-\subsubsection{Safety}
-
-Type safety for $\lambdap$ is straight-forward, but is necessary in order to establish the 
+Type preservation for $\lambdap$ is essentially trivial, but is necessary in order to establish the 
 correctness of our translation.
-Again, we elide a more careful treatment by treating a special case where our language terminates per the discussion above.
+Again, we give a small step semantics and address type safety more completely in \cite{psptr}.
 
 \begin{thm}[Safety for $\lambdap$] If $\emptyset \vdash \iota : \tau$ 
-  then $\iota \sreduces \dot v$ and $\emptyset \vdash \dot v : \tau$.
+  and $\iota \sreduces \dot v$ then $\emptyset \vdash \dot v : \tau$.
 \end{thm}
@@
 \begin{lem}[Canonical Forms for Target Language]
-~
+If $\emptyset \vdash \dot{v} : \tau$ then 
 \begin{enumerate}
 \item 
-If $\emptyset \vdash \iota : \regex$ then $\iota \treduces \rx{r}$ such that $r$ is a well-formed regular expression. 
+If $\tau=\regex$ then $\dot{v}= \rx{r}$ such that $r$ is a well-formed regular expression. 
 \item 
-If $\emptyset \vdash \iota : \str$ then $\iota \treduces \tstr{s}$.
+If $\tau=\str$ then $\dot{v} = \tstr{s}$.
 \end{enumerate}
@@
 The coercion cases are most interesting. If the safety of coercion in manifest in the
-types of the expressions, then no runtime check is inserted.
-If the safety of coercion is not manifest in the types, then a check is inserted. Note that regular strings translate to strings directly; there is no space overhead.
+types of the expressions, then no runtime check is inserted (\textsc{Tr-SafeCoerce}).
+If the safety of coercion is not manifest in the type, then a check is inserted (\textsc{Tr-Check}). Note that regular strings translate to strings directly; there is no space overhead.
 
@@
   and $\trden{\tctx} \vdash \iota : \trden{\sigma}$.
-  Furthermore, $e \sreduces v$ and
+  Furthermore, if $e \sreduces v$ then
   $\iota \treduces \dot{v}$ such that
@@
 \begin{thm}[Correctness of Input Sanitation for Translated Terms]\label{thm:main}
-  If $\trden{e} = \iota$ and $\emptyset \vdash e : \stringin{r}$ then $\iota \sreduces \tstr{s}$
+  If $\trden{e} = \iota$ and $\emptyset \vdash e : \stringin{r}$ and $e \sreduces \rstr{s}$ then $\iota \sreduces \tstr{s}$
   for $s \in \lang{r}$.
@@
 \begin{prf}
-  By Theorem \ref{thm:trcorrect} and the rules given, $\iota \sreduces \tstr{s}$ implies that $e \sreduces \strin{s}$.
-  Theorem \ref{thm:scorrect} together with the assumption that $e$ is well-typed implies that $s \in \lang{r}$.
+  By Theorem \ref{thm:trcorrect}, we have our first conclusion. By Theorem \ref{thm:scorrect} together with the assumption that $e$ is well-typed we have that $s \in \lang{r}$.
 \end{prf}
@@
 
-\inferrule[Tr-Subst]
+\inferrule[Tr-Replace]
 {   \trden{e_1} = \iota_1 \\   \trden{e_2} = \iota_2 }
@@
 \bibliographystyle{abbrv}
-\renewcommand*{\bibfont}{\footnotesize}
+%\renewcommand*{\bibfont}{\small}
 \setlength{\bibsep}{0pt plus .25pt minus .25pt}

### 2014-12-13 sanitation-psp14/tr/strings_chapter.tex
Saturday/sunday.

@@
 \renewcommand{\thetr}{\Alph{tr}} % "letter-numbered" theorems
-\newtheorem{trexam}[tr]{TR Example}
-\newtheorem{trthm}[tr]{TR Theorem}
-\newtheorem{trlem}[tr]{TR Lemma}
-\newtheorem{trdef}[tr]{TR Definition}
-\newtheorem{trconj}[tr]{TR Conjecture}
+\newtheorem{trexam}[tr]{Example}
+\newtheorem{trthm}[tr]{Theorem}
+\newtheorem{trlem}[tr]{Lemma}
+\newtheorem{trdef}[tr]{Definition}
+\newtheorem{trconj}[tr]{Conjecture}
+\newtheorem{ass}[tr]{Assumption}
 
@@
 
+Throughout the following proofs, we abbreviate $s \in \lang{r}$ by $s \in r$.
+
 \begin{trlem}[Leading characters are in the head] \label{trlemmax}
-If $c_1 \cdot c_2 \cdot ... \cdot c_m \in \lang{r}$, then
+If $s = c_1  c_2 \cdots c_m \in \lang{r}$, then
 $c_1 \in \lhead{r}$.
@@
 \begin{proof}
-By structural induction on $r$...
+By structural induction on $r$.
+
+If $r = x$ for some character $x$ then $s \in \lang{r}$ implies $s = x$.
+By definition of lhead, $\lhead{s} = \lhead{s, \epsilon} = \lhead{x, \epsilon} = x$.
+Finally, $x \in \lang{x}$.
+
+If $r = r_1 \cdot r_2$ then $s \in \lang{r}$ implies $s = c_1 s_1 s_2$
+such that $c_1 s_1 \in \lang{r_1}$. 
+By definition of lhead, $\lhead{r_1 \cdot r_2} = \lhead{r_1, r_2} =
+\cdots + \lhead{r_1, \epsilon} = \cdots + \lhead{r_1}$. Therefore, $r_1 \subseteq \lhead{r_1, r_2}$
+and so it suffices to show that $s = c_1 s_1 s_2$ such that $c_1 \in \lhead{r_1}$,
+\todo{this is why we break out the $c_1$ from the $s_1$, both here and going forward}
+which holds by induction because $c_1 s_1 \in \lang{r_1}$.
+
+If $r = r_1 + r_2$ then $s \in r$ implies either $s \in r_1$ or $s \in r_2$.
+Without loss of generality suppose $s \in r_1$.
+Then by induction $s = c_1 \cdots c_m$ such that $c_1 \in \lhead{r_1}$.
+By definition, $\lhead{r} = \lhead{r_1, \epsilon} + \lhead{r_2, \epsilon} = \lhead{r_1} + \cdots$.
+Therefore, $c_1 \in \lhead{r}$.
+
+If $r = r_1^*$ then $s \in r$ implies that $s = s_1 \cdots s_k$ where $k \ge 0$ and $s_i \in \lang{r}$.
+By definition of lhead, $\lhead{r_1^*} = \epsilon + \lhead{r_1}$.
+If $s = \epsilon$, then $s \in \lang{\epsilon} \subseteq \lang{\lhead{r_1}}$.
+If $s = s_1 \cdots s_k$, then $s_1 \in \lang{r}$ and so by induction $s_1 = c_1 \cdots c_k$ such that $c_1 \in \lang{\lhead{r}}$.
 \end{proof}
@@
 \begin{trthm}[Correctness of \textsf{ltail}]
-If $s \in \lang{r}$ then $s \in \lang{\lhead{r}} \cdot \lang{\ltail{r}}$.
+If $s \in \lang{r}$ then $s \in \lang{\lhead{r} \cdot \ltail{r}}$.
 \end{trthm}
@@
 The proof proceeds by structural induction on $r$.
-In each case, we demonstrate three facts:
-\begin{itemize}
-  \item First, that the string $s = a \cdot b$ is a concatenation of two substrings (each of which may be $\epsilon$).
-  \item Second, $a \in \lang{\lhead{r}}$ (or, equivalently, $s \in \lang{\lhead{r}}$ letting $b = \epsilon$).
-  \item Third, there exists a $c \in \lhead{r}$ such that $a = c \cdot a'$ and $a' \cdot b \in \lang{\delta_c(r)}$, or else $s = a$. 
-\end{itemize}
-
-Throughout, we consider $\epsilon$ as idenetity under all operators for expressions and $\cdot$ for strings.
+In each case, we identify sublanguages of the head and tail of $r$ such that 
+$s$ is in the concatenation of the the sublanguages.
+Throughout the proof, we use the fact that $\lang{r'} \subseteq \lang{r}$ and $\lang{q'} \subseteq \lang{q}$
+implies $\lang{r' \cdot q'} \subseteq \lang{r \cdot q}$.
 
@@
 \item \textbf{Case $r = c$ for $c \in Sigma$}.
-Note that if $s \in \lang{c}$ then $s = c$.
-Note that $\lhead{c} = \lhead{c, \epsilon} = c$, and $c \in \lang{c}$.
-So $c \in \lang{c}$, which is sufficient since $\epsilon \in \ltail{r}$.
+By definition, if $s \in \lang{c}$ then $s = c$.
+By definition of lhead and ltail, $\lhead{c} = \lhead{c, \epsilon} = c$
+and $\lang{\epsilon} \subseteq \lang{\ltail{r}}$.
+Therefore, $c \in \lang{c \cdot \epsilon} \subseteq \lang{\lhead{r} + \ltail{r}}$.
 
@@
 Suppose $s \in \lang{r}$.
-We may decompose $s$ as $s_1 \cdot ... \cdot s_n$ such that $s_1 = c_1 \cdot ... \cdot c_m \in r_1$.
+We may decompose $s$ as $s_1  \cdots s_n$ such that $s_1 = c_1 s_1' \in \lang{r_1}$.
 Note that $c_1 \in \lhead{r_1}$ by \ref{trlemmax}.
-Let $a' = c_2 \cdot ... \cdot c_m$.
-It suffices to show that $a' \cdot b \in \lang{\delta_c(r)}$.
-Since $s = c_1 \cdot a' \cdot b$ and $s \in \lang{r}$, it follows by the definition of derivative that $a' \cdot b \in \delta_{c_1}(r)$.
+Therefore, $c_1 \in \lhead{r_1 \cdot r_2}$ by definition of lhead.
+It suffices to show that $s_1' s_2 \in \lang{\delta_{c_1}(r)}$.
+Since $s = c_1 s_1' s_2\in \lang{r}$, it follows by the definition of derivative that $s_1' s_2 \in \delta_{c_1}(r)$.
+Therefore, $s = c_1 s_1' s_2 \in \lang{ c_1 \cdot \delta_{c_1} } \subseteq \lang{ \lhead{r} \cdot \ltail{r}}$.
 
@@
 so that $s \in \lang{r_1}$ or $s \in \lang{r_2}$.
-In either case, the result follows by induction.
+Consider without loss of generality the case where $s \in \lang{r_1}$.
+By induction, $s \in \lang{\lhead{r_1} \cdot \ltail{r_1}}$
+Note that $\lhead{r_1} \subseteq \lhead{r}$
+and $\ltail{r_1} \subseteq \ltail{r}$.
+Therefore, $\lhead{r_1} \cdot \ltail{r_2} \subseteq \lhead{r_1} \cdot \ltail{r_1}$.
 
 \item \textbf{Case $r = q^*$}.
-Note that $s = \epsilon$ or else $s = s_1 \cdot s_2 \cdot ... s_n$ where each $s_i \in \lang{q}$\footnote{intuitively, $s$ is either empty or a finitary concatenation of strings matching $q$.}.
+Either $s$ is the empty string or else the $k^{th}$ unwinding (i.e., $s = s_1 \cdots s_k$ where $k > 0$ and $s_i \in \lang{q}$).
 
 Suppose $s = \epsilon$.
-Note that $\lhead{r} = \lhead{q^*} = \lhead{q^*, \epsilon} = \lhead{\epsilon, \epsilon} + h' = \epsilon + h'$.
-Therefore, $s \in \lhead{r}$.
-
-In the other case, suppose $s = s_1 \cdot ... \cdot s_n$ where each $s_i \in \lang{q}$.
-Let $a = s_1$ and $b = s_2 \cdot ... s_n$, so that $s = a \cdot b$.
-Note that $a = c_1 \cdot ... \cdot c_n$.
-We will show that $c_1 \in \lang{\lhead{q^*}}$ and that $(c_2 \cdot ... c_n) \cdot b \in \delta_{c_1}(q^*)$.
-
-The latter property is trivial. Note that $s = c_1 \cdot c_2 \cdot ... \cdot c_n \cdot b \in \lang{q^*}$.
-The result follows immediately from the definition of derivative.
-
-What remains to be shown is that $c_1 \in \lang{q^*}$.
-Note that $\lhead{q^*} = \epsilon + \lhead{q, \epsilon}$.
-Recall that $a = s_1$ where $s_1 \in \lang{q}$.
-Also recall that $a= c_1 \cdot ... \cdot c_n$.
-Therefore, $c_1 \in \lhead{q}$ by \ref{trlemmax}.
-
+By definition of lhead, $\lhead{r} = \lhead{q^*} = \lhead{q^*, \epsilon} = \lhead{\epsilon, \epsilon} + \cdots = \epsilon + \cdots$.
+By definition of ltail, $\ltail{r} = \epsilon \cdots$.
+From these facts we conclude that $s = \epsilon \in \lang{\epsilon \cdot \epsilon} \subseteq \lang{\lhead{r} \cdot \ltail{r}}$.
+
+Suppose instead that $s = s_1 \cdots s_k$ for $k < 0$ where $s_i \in \lang{q}$.
+If $q = \epsilon$ then $s = \epsilon$ and the conclusion follows for the same reason as the epsilon subcase of the character case.
+Otherwise, $s$ is not the empty string and so $s = c_1 s_1 s_2$ where $s_1$ and $s_2$ might be empty.
+By the lemma above, $c_1 \in \lhead{s}$.
+By the definition of derivative, $s_1s_2 \in \delta_{c_1}(s)$.
+Therefore, $s \in \lang{\lhead{s} \cdot \delta_{c_1}(s)} \subseteq \lang{\lhead{r} \cdot \ltail{r}}$, where the set inclusion follows by the definition
+of lhead and ltail.
 \end{itemize}
@@
 
-\begin{trexam}[All the heads of all the tails can be more than one head and tail]\label{exam:headsandtails}
+\begin{trexam}[Decomposition is not Reversible]\label{exam:headsandtails}
 $r \not = \lhead{r} \cdot \ltail{r}$.
@@
 A simple counter-example is $ab + cd$. 
-Note that $\lhead{ab + cd} = a+c$ and $\ltail{ab + cd} = b + d$. 
-Therefore, $\{ad,bc\} \subset \lang{\lhead{ab + cd} \cdot \ltail{ab + cd}}$ even though neither of these is in $\lang{r}$.
+Note that $\lhead{a\cdot b + c\cdot d} = a+c$ and $\ltail{a\cdot b + c\cdot d} = b + d$. 
+Therefore, $\{ad,bc\} \subset \lang{\lhead{a\cdot b + cd\cdot } \cdot \ltail{a\cdot b + c\cdot d}}$ even though neither of these is in $\lang{r}$.
 \end{proof}
@@
 property required for soundness.
-Still, in a production implementation, it will make sense to massage the definitions of $\lhead{r}$ and $\ltail{r}$ so that type information is not unnecessarily lost during substring operations.
 
@@
 the effect of a substitution. Closing these gaps in approximation is important, and motivates the string operations portion of this technical report.
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+% Substitution.
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 
@@
 
-\begin{lem}[Properties of Regular Languages.] \label{thm:regexprops}
+\begin{ass}[Properties of Regular Languages.] \label{thm:regexprops}
 ~
-\begin{itemize}
+\begin{enumerate}
 
@@
 \item Regular languages are closed under reversal.
-\end{itemize}
-\end{lem}
+\end{enumerate}
+\end{ass}
 
@@
 
-\begin{lem}
-  If $\emptyset \vdash e : \stringin{r}$ then $r$ is a well-formed regular expression.
-\end{lem}
-\begin{proof}
-  The proof proceeds by induction on the derivation of the premise.
-  The only non-trivial cases (those which require more than an appeal to inversion) are S-T-Case, S-T-Replace and S-T-Concat.
-
-In the S-T-Case case, note that ${\sf lhead}$ and ${\sf ltail}$ are total functions for well-formed regular expressions to well-formed regular expressions.
-
-In the S-T-Concat case, note that Lemma~\ref{thm:total} implies that if $r_1$ and $r_2$ are regular expressions then so is $r_1 \cdot r_2$.
-
-In the S-T-Replace case, it suffices to show that $\lreplace{r}{r_1}{r_2}$ is a regular expression assuming (inductively) that $r, r_1$ and $r_2$ are all regular expressions.
-This follows from the Closure proposition.
-\end{proof}
+%\begin{lem}
+%  If $\emptyset \vdash e : \stringin{r}$ then $r$ is a well-formed regular expression.
+%\end{lem}
+%\begin{proof}
+%  The proof proceeds by induction on the derivation of the premise.
+%  The only non-trivial cases (those which require more than an appeal to inversion) are S-T-Case, S-T-Replace and S-T-Concat. 
+%
+%In the S-T-Case case, note that ${\sf lhead}$ and ${\sf ltail}$ are total functions for well-formed regular expressions to well-formed regular expressions.
+%
+%In the S-T-Concat case, note that Lemma~\ref{thm:total} implies that if $r_1$ and $r_2$ are regular expressions then so is $r_1 \cdot r_2$.
+%
+%In the S-T-Replace case, it suffices to show that $\lreplace{r}{r_1}{r_2}$ is a regular expression assuming (inductively) that $r, r_1$ and $r_2$ are all regular expressions.
+%This follows from the Closure proposition.
+%\end{proof}
 
@@
 
-\begin{trconj}
-if $\emptyset \vdash e : \sigma$ then $e \mapsto^* v$ such that $v \ \val$.
-\end{trconj}
-%\begin{trthm}[Termination of small step semantics]
-%\end{trthm}
-%\begin{proof}
-We do not develop the full proof here, but note that the simply typed lambda calculus terminates.
-For the string fragment, observe that the S-T- rules do not add any non-trivial binding structure because substitutions
-$[e/x]e'$ may only occur in the special case where $e = \rstr{s}$, so that the length of the term
-never increases and the number of free variables strictly decreases. Therefore, the standard normalization
-argument proceeds without complication after fixing an evaluation order for the compatibility rules 
-(all our other proofs are agnostic to evaluation order).
+%\begin{trconj}
+%if $\Gamma \vdash e : \sigma$ then $e \mapsto^* v$ such that $v \ \val$.
+%\end{trconj}
+%%\begin{trthm}[Termination of small step semantics]
+%%\end{trthm}
+%%\begin{proof}
+%We do not develop the full proof here, but note that the simply typed lambda calculus terminates.
+%For the string fragment, observe that the S-T- rules do not add any non-trivial binding structure because substitutions
+%$[e/x]e'$ may only occur in the special case where $e = \rstr{s}$, so that the length of the term
+%never increases and the number of free variables strictly decreases. Therefore, the standard normalization
+%argument proceeds without complication after fixing an evaluation order for the compatibility rules 
+%(all our other proofs are agnostic to evaluation order).
 %\end{proof}
@@
 \begin{trlem}[Canonical Forms]\label{cfgen}
-Suppose $v \ \val$.%or $e \sreduces v$.
-
-If $\emptyset\vdash v : \stringin{r}$ then $v = \rstr{s}$.
-
-If $\emptyset\vdash v : \sigma \rightarrow \sigma'$ then $v = \lambda x . e'$ for some $e'$.
+If $\emptyset \vdash v : \sigma$ then:
+\begin{enumerate}
+\item If $\sigma = \stringin{r}$ then $u = \rstr{s}$.
+\item If $\sigma = \sigma_1 \rightarrow \sigma_2$ then $v = \lambda x . e'$ and $s \in \lang{r}$.
+\end{enumerate}
+
 \end{trlem}
@@
 \begin{proof}
-This fact follows directly from Lemma~\ref{cfgen}.
+This fact follows directly from Lemma~\ref{cfgen} and the preservation lemma.
 \end{proof}
@@
 \begin{trlem}[Progress of small step semantics.]
-If $\emptyset\vdash e : \sigma$ either $e \ \val$ or $e \mapsto e'$ for some $e'$.
+If $\emptyset\vdash e : \sigma$ either $e = v$ or $e \mapsto e'$ for some $e'$.
 \end{trlem}
@@
 \begin{itemize}[label=$ $,itemsep=1ex]
-\item \textbf{$\lambda$ fragment}. Cases SS-T-Var, SS-T-Abs, and SS-T-App are exactly as in a proof of type safety for
+\item \textbf{$\lambda$ fragment}. Cases SS-T-Var, SS-T-Abs, and SS-T-App are exactly as in a proof of progress for
 the simply typed lambda calculus.
@@
 \item \textbf{S-T-Stringin-I}. Suppose $\emptyset\vdash \rstr{s} : \stringin{s}$.
-The $\rstr{s} \ \val$ by SS-E-RStr.
-
-\item \textbf{S-T-Concat}. Suppose $\emptyset\vdash \rconcat{e_1}{e_2} : \stringin{s}$.
-By inversion and induction, $e_1 \mapsto e_1'$ or $e_1 \ \val$ and similarly
+Then $\rstr{s} \ \val$ by SS-E-RStr.
+
+\item \textbf{S-T-Concat}. Suppose $\emptyset\vdash \rconcat{e_1}{e_2} : \stringin{r}$ and $\emptyset e_1 : \stringin{r_1}$ and $\emptyset \vdash e_2 : \stringin{r_2}$.
+By induction, $e_1 \mapsto e_1'$ or $e_1 = v_1$ and similarly
 for $e_2$. 
@@
 
-In the remaining case, $e_1 \ \val$ and $e_2 \ \val$.
+In the remaining case, $e_1 =v_1$ and $e_2 =v_2$.
 But then it follows by Canonical Forms that $e_1 = \rstr{s_1}$
 and $e_2 = \rstr{s_2}$.
-Finally, by SS-E-Concat, \\ $\rconcat{\rstr{s_1}}{\rstr{s_2}} \mapsto \rstr{s_1s_2}$.
+Finally, by SS-E-Concat, $\rconcat{\rstr{s_1}}{\rstr{s_2}} \mapsto \rstr{s_1s_2}$.
 
@@
 \item \textbf{S-T-Case}.
-Suppose $e = \strcase{e_1}{e_2}{x,y.e_3}$. 
-By inversion,\\ $\emptyset\vdash e_1 : \stringin{r}$.
+Suppose $e = \strcase{e_1}{e_2}{x,y.e_3}$ and $\emptyset\vdash e_1 : \stringin{r}$.
 %\begin{align}
@@
 %In the former case, $e \mapsto \strcase{e_1}{e_2'}{x,y.e_3}$.
-From this fact, induction, and Canonical Forms it follows that $e_1 \mapsto e_1'$ or $e_1 = \rstr{s}$.
+By induction and Lemma~\ref{thm:cfs} it follows that $e_1 \mapsto e_1'$ or $e_1 = \rstr{s}$.
 In the former case, $e$ steps by S-E-Case-Left.
@@
 \item \textbf{S-T-Replace}.
-Suppose $e = \rreplace{r}{e_1}{e_2}$ and $\emptyset\vdash e : \stringin{r'}$ where, by inversion of S-T-Replace,
+Suppose $e = \rreplace{r}{e_1}{e_2}$, $\emptyset\vdash e : \stringin{r'}$ and:
 \begin{align}
@@
 
-By (1), inversion and induction $e_1 \ \val$ or $e_1 \mapsto e_1'$ for some $e_1'$,
+By induction on (1), $e_1 \mapsto e_1'$ or $e_1 = v_1$  for some $e_1'$.
 If $e_1 \mapsto e_1'$ then $e$ steps by SS-E-Replace-Left.
 Similarly, if $e_2$ steps then $e$ steps by SS-E-Replace-Right.
-The only remaining case is where $e_1 \ \val$ and also $e_2 \ \val$.
-But then by Canonical Forms, $e_1 = \strin{s_1}$ and $e_2 = \strin{s_2}$.
+The only remaining case is where $e_1 = v_1$ and also $e_2 = v_2$.
+By Canonical Forms, $e_1 = \strin{s_1}$ and $e_2 = \strin{s_2}$.
 Therefore, $e \mapsto \rstr{\lsubst{r}{s_1}{s_2}}$ by SS-E-Replace.
@@
 Suppose that $\emptyset\vdash \rcoerce{r}{e_1} : \stringin{r}$.
-By inversion of S-T-SafeCoerce,
+and
 %\begin{align}
@@
 %\end{align}
-By induction, $e_1 \ \val$ or $e_1 \mapsto e_1'$ for some $e_1'$.
+By induction, $e_1 =v_1$ or $e_1 \mapsto e_1'$ for some $e_1'$.
 If $e_1 \mapsto e_1'$ then $e$ steps by SS-E-SafeCoerce-Step.
-Otherwise, $e_1 \ \val$ and by Canonical Forms $e_1 = \rstr{s}$.
+Otherwise, $e_1 =v$ and by Canonical Forms $e_1 = \rstr{s}$.
 In this case, $e = \rcoerce{r}{\rstr{s}} \mapsto \rstr{s}$ by SS-E-SafeCoerce.
@@
 \item \textbf{S-T-SafeCheck}
-Suppose that $\emptyset\vdash \rcheck{r}{e_0}{x.e_1}{e_2} : \stringin{r}$.
-By inversion of S-T-Check:
+Suppose that $\emptyset\vdash \rcheck{r}{e_0}{x.e_1}{e_2} : \stringin{r}$ and:
 \begin{align}
@@
 
-By (\ref{align6}) and induction,  $e_0 \mapsto e_0'$ or $e_0 \ \val$.
+By induction,  $e_0 \mapsto e_0'$ or $e_0 = v$.
 In the former case $e$ steps by SS-E-Check-StepRight.
 Otherwise, $e_0 = \rstr{s}$ by Canonical Forms.
-By Lemma~\ref{thm:regexprops}, either $s \in \lang{r_0}$ or $s \not\in\lang{r_0}$.
+By Lemma~\ref{thm:regexprops} part 2, either $s \in \lang{r_0}$ or $s \not\in\lang{r_0}$.
 In the former case $e$ takes a step by SS-E-Check-Ok.
@@
 
+\begin{trlem}[Substitution] \label{lem:subst}
+If $\Psi\ x : \sigma \vdash e : \sigma'$ and $\psi \vdash e' : \sigma$, then
+$\psi \vdash [e' / x] e : \sigma'$.
+\end{trlem}
+
 \begin{trlem}[Preservation for Small Step Semantics]
-If $\emptyset \vdash e : \sigma$ and $e \mapsto e'$ then $\emptyset \vdash e : \sigma$.
+If $\emptyset \vdash e : \sigma$ and $e \mapsto e'$ then $\emptyset \vdash e' : \sigma$.
 \end{trlem}
@@
 
-By induction on the derivation of $e \mapsto e'$.
+By induction on the derivation of $e \mapsto e'$ and $\emptyset \vdash e : \sigma$.
 
@@
 Suppose $e = \rconcat{e_1}{e_2} \mapsto \rconcat{e_1'}{e_2}$ and $e_1 \mapsto e_1'$.
-By inversion of S-T-Concat, $\emptyset\vdash e_1 : \stringin{r_1}$ where \\$\emptyset\vdash e : \stringin{r_1r_2}$.
-By induction, if $e_1 \mapsto e_1'$ then $\emptyset\vdash e_1' : \stringin{r_1}$.
-Therefore, $\emptyset\vdash \rconcat{e_1'}{e_2} : \stringin{r_1r_2}$.
-
-\item {\textbf{SS-E-Concat-Right}}. \todo{Similar to SS-E-Concat-Left.}
-
-\item \textbf{SS-E-Concat}. Suppose $\emptyset\vdash\rconcat{\rstr{s_1}}{\rstr{s_2}} : \stringin{r_1r_2}$ and
-$\rconcat{\rstr{s_1}}{\rstr{s_2}} \mapsto \rstr{s_1s_2}$.
-Then by inversion $\emptyset\vdash\rstr{s_1} : \stringin{r_1}$ and similarly for $\rstr{s_2}$.
-Therefore, $s_1 \in \lang{r_1}$ and\\ $s_2 \in  \lang{r_2}$ from which it follows by Lemma~\ref{thm:regexprops}
+The only rule that applies is S-T-Concat, so $\emptyset\vdash e_1 : \stringin{r_1}$ and $\emptyset\vdash e_2 : \stringin{r_2}$.
+By induction, $\emptyset\vdash e_1' : \stringin{r_1}$.
+Therefore, by S-T-Concat, $\emptyset\vdash \rconcat{e_1'}{e_2} : \stringin{r_1r_2}$.
+
+\item {\textbf{SS-E-Concat-Right}}. 
+Suppose $e = \rconcat{e_1}{e_2} \mapsto \rconcat{e_1}{e_2'}$ and $e_2 \mapsto e_2'$.
+The only rule that applies is S-T-Concat, so $\emptyset\vdash e_1 : \stringin{r_1}$ and $\emptyset\vdash e_2 : \stringin{r_2}$.
+By induction, $\emptyset\vdash e_2' : \stringin{r_2}$.
+Therefore, by S-T-Concat, $\emptyset\vdash \rconcat{e_1}{e_2'} : \stringin{r_1r_2}$.
+
+\item \textbf{SS-E-Concat}. 
+Suppose $\rconcat{\rstr{s_1}}{\rstr{s_2}} \mapsto \rstr{s_1s_2}$.
+The only applicable rule is S-T-Concat, so 
+$\emptyset\vdash\rstr{s_1} : \stringin{r_1}$ 
+and
+$\emptyset\vdash\rstr{s_2} : \stringin{r_2}$. 
+By Lemma~\ref{lem:cfs}, $s_1 \in \lang{r_1}$ and $s_2 \in  \lang{r_2}$ from which it follows by Lemma~\ref{thm:regexprops}
 that $s_1s_2 \in \lang{r_1r_2}$. Therefore, $\emptyset\vdash \rstr{s_1s_2} : \stringin{r_1r_2}$ by S-T-Rstr.
@@
 \item \textbf{S-E-Case-Left}.
-Suppose that $e = \strcase{e_1}{e_2}{x,y.e_3}$ and also that
+Suppose 
 $e \mapsto \strcase{e_1'}{e_2}{x,y.e_3}$
-and $\emptyset \vdash e : \stringin{r}$.
-By inversion of S-T-Case:
+and $\emptyset \vdash e : \sigma$ and $e_1 \rightarrow e_1'$.
+The only rule that applies is S-T-Case, so:
 \begin{align}
@@
 
-\item \textbf{SS-E-Case-Right}. 
-We have that $e = \strcase{e_1}{e_2}{x,y.e_3}$, Suppose $e \mapsto \strcase{e_1}{e_2'}{x,y.e_3}$
-and $\emptyset \vdash e : \stringin{r}$.
-By inversion of S-T-Case:
-\begin{align}
-\emptyset&\vdash e_1 : \stringin{r} \label{align10}\\
-\emptyset&\vdash e_2 : \sigma \label{align11} \\ 
-x:\stringin{\lhead{r}}, y : \stringin{\ltail{r}} &\vdash e_3 : \sigma \label{align12}
-\end{align}
-By (\ref{align11}) and the assumption that $e_2 \mapsto e_2'$, it follows by induction that $\emptyset\vdash e_2' : \stringin{r}$.
-This fact together with (\ref{align10}) and (\ref{align12}) implies by S-T-Case that 
-$\emptyset\vdash\strcase{e_1}{e_2'}{x,y.e_3} : \sigma$.
+%\item \textbf{SS-E-Case-Right}. 
+%We have that $e = \strcase{e_1}{e_2}{x,y.e_3}$, Suppose $e \mapsto \strcase{e_1}{e_2'}{x,y.e_3}$
+%and $\emptyset \vdash e : \stringin{r}$.
+%By inversion of S-T-Case:
+%\begin{align}
+%\emptyset&\vdash e_1 : \stringin{r} \label{align10}\\
+%\emptyset&\vdash e_2 : \sigma \label{align11} \\ 
+%x:\stringin{\lhead{r}}, y : \stringin{\ltail{r}} &\vdash e_3 : \sigma \label{align12}
+%\end{align}
+%By (\ref{align11}) and the assumption that $e_2 \mapsto e_2'$, it follows by induction that $\emptyset\vdash e_2' : \stringin{r}$.
+%This fact together with (\ref{align10}) and (\ref{align12}) implies by S-T-Case that 
+%$\emptyset\vdash\strcase{e_1}{e_2'}{x,y.e_3} : \sigma$.
 
@@
 Suppose: \begin{align*}
-&e = \strcase{-}{e_2}{-} \\
+&e = \strcase{e_0}{e_2}{x,y.e_3} \\
 &\emptyset\vdash e: \sigma \\
@@
 \end{align*}
-By inversion of S-T-Case, $e_2 : \sigma$.
+The only rule htat applies is S-T-Case, so $e_2 : \sigma$.
 
@@
 \emptyset&\vdash \rstr{as} : \stringin{r} \label{align13}\\
-\emptyset&\vdash \rstr{e_2} : \sigma \label{align14} \\ 
+\emptyset&\vdash e_2 : \sigma \label{align14} \\ 
 x:\stringin{\lhead{r}}, y : \stringin{\ltail{r}} &\vdash e_3 : \sigma \label{align15}
@@
 
-We know that $as \in \lang{r}$ by (\ref{align13}) and inversion of S-T-Rstr.
+We know that $as \in \lang{r}$ by canonical forms on (\ref{align13}) 
 Therefore, $a \in \lang{\lhead{r}}$ by definition of ${\sf lhead}$.
-Furthermore, $\ltail{r} = ... | \delta_a{r} | ...$ by definition of ${\sf ltail}$.
+Furthermore, $\ltail{r} = ... + \delta_a{r} + ...$ by definition of ${\sf ltail}$.
 Note that $s \in \lang{\delta_a{r}}$ by definition of the derivative, and so
@@
 and $\emptyset\vdash \rstr{s} : \stringin{\lhead{r}}$.
-It follows by (\ref{align15}) that  $\emptyset\vdash [\rstr{a}, \rstr{s} / x,y] e_3 : \sigma$.
-
-\item \textbf{Cases SS-E-Replace-Left, SS-E-Replace-Right, SS-E-Check-StepLeft, SS-E-SafeCoerce-Step, SS-E-Check-StepRight}.
-\todo{At this point the method for handling compatibility cases is clear; therefore,
-we elide these cases.}
+It follows by (\ref{lem:subst}) that  $\emptyset\vdash [\rstr{a}, \rstr{s} / x,y] e_3 : \sigma$.
+
+\item \textbf{Cases SS-E-Replace-Left, SS-E-Replace-Right, SS-E-Check-StepLeft, SS-E-SafeCoerce-Step, SS-E-Check-StepRight}. \todo{todo}
 
@@
 Suppose $e = \rreplace{r}{\rstr{s_1}}{\rstr{s_2}} \mapsto \rstr{\lsubst{r}{s_1}{s_2}}$.
-Assume $\emptyset \vdash e : \stringin{r'}$ for $r' = \lreplace{r}{r_1}{r_2}$.
-Then by inversion of S-T-Replace:
+The only applicable rule is 
+S-T-Replace, so
 \begin{align*}
@@
 \end{align*}
-from which follows that $s_1 \in \lang{r_1}$ and $s_2 \in \lang{r_2}$.
-Therefore, $\lsubst{r}{s_1}{s_2} \in \lang{\lreplace{r}{r_1}{r_2}}$ by Theorem~\ref{thm:substcorrespondence}.
+By conanical forms,
+$s_1 \in \lang{r_1}$ and $s_2 \in \lang{r_2}$.
+Therefore, $\lreplace{r}{s_1}{s_2} \in \lang{\lreplace{r}{r_1}{r_2}}$ by Theorem~\ref{thm:substcorrespondence}.
 It is finally derivable by S-T-Rstr that:
 
-$\emptyset \vdash \rstr{\lsubst{r}{s_1}{s_2}} : \stringin{\lreplace{r}{r_1}{r_2}}$.
+$\emptyset \vdash \rstr{\lreplace{r}{s_1}{s_2}} : \stringin{\lreplace{r}{r_1}{r_2}}$.
 
 \item \textbf{Case SS-E-SafeCoerce}.
-Suppose that $\rcoerce{r}{s_1} \mapsto \rstr{s_1}$ and that $\emptyset\vdash \rcoerce{r}{s_1} : \stringin{r}$.
-By inversion of S-T-SafeCoerce we know that $s \in \lang{r}$. Therefore, $\emptyset\vdash s : \stringin{r}$.
+Suppose that $\rcoerce{r}{\rstr{s_1}} \mapsto \rstr{s_1}$.
+The only applicable rule is S-T-SafeCoerce, so 
+$\emptyset\vdash \rcoerce{r}{s_1} : \stringin{r}$.
+By Canonical Forms, $s \in \lang{r}$. Therefore, $\emptyset\vdash \rstr{s} : \stringin{r}$.
 
@@
 $s \not \in \lang{r}$, and $\emptyset \vdash \rcheck{r}{\rstr{s}}{x.e_1}{e_2} : \sigma$.
-By inversion of S-T-Check, $\emptyset \vdash e_2 : \sigma$.
+The only applicable rule is S-T-Check, so $\emptyset \vdash e_2 : \sigma$.
 
@@
 The only applicable rule is S-E-Abs, so $v = \lambda x . e_1$.
-Note that $\lambda x . e_2 \mapsto^* \lambda x . e_2$ by RT-Refl.
+Note that $\lambda x . e_1 \mapsto^* \lambda x . e_1$ by RT-Refl.
 
@@
   If $\emptyset \vdash e : \sigma$ 
-  then $e \sreduces v$ and $\emptyset \vdash v : \sigma$.
+  and $e \sreduces e'$ then $\emptyset \vdash e' : \sigma$.
 \end{thm}
 \begin{proof}
-If $\emptyset \vdash e : \sigma$ then $e \mapsto^* v$ for $v \ \val$ by termination.
-Therefore, $e \sreduces v$ by part 2 of the semantic correspondence theorem.
-
-Since $\emptyset \vdash e : \sigma$ and $e \mapsto^* v$, it follows that $\emptyset \vdash v : \sigma$ by type safety for the small step semantics.
+If $\emptyset \vdash e : \sigma$ then $e \mapsto^* e'$.
+Therefore, $e \sreduces e'$ by part 2 of the semantic correspondence theorem.
+
+Since $\emptyset \vdash e : \sigma$ and $e \mapsto^* e'$, it follows that $\emptyset \vdash e' : \sigma$ by type safety for the small step semantics.
 \end{proof}
@@
 If $\emptyset \vdash e : \stringin{r}$ and $e \sreduces \rstr{s}$ then $\emptyset \vdash \rstr{s} : \stringin{r}$ by Type Safety.
-By inversion of S-T-Rstr, $s \in \lang{r}$.
+By Caonical Forms, $s \in \lang{r}$.
 \end{proof}
@@
 
-Choose $\iota = \tconcat{\iota_1}{\iota_2}{x,y.\iota_3}$ \todo{and note that by the properties established via induction,
-$\trden{e} = \iota$ and $\trden{\sctx} \vdash \iota : \trden{\sigma}$.}
+Choose $\iota = \tconcat{\iota_1}{\iota_2}{x,y.\iota_3}$ and note that by the properties established via induction,
+$\trden{e} = \iota$ and $\trden{\sctx} \vdash \iota : \trden{\sigma}$.
 

### 2014-12-13 sanitation-psp14/sanitation-psp14.tex

--- 2014-08-05 sanitation-psp14/sanitation-psp14.tex
@@
-\documentclass{acm_proc_article-sp}
+\documentclass[10pt]{sigplanconf}
 \newcommand{\lamAce}{\lambda_{\text{Ace}}}
@@
     \makeatletter
-\usepackage{enumitem}
-\usepackage{enumerate}
+%\usepackage{enumitem}
+\usepackage{enumerate}%
 
@@
     numbers=left,
-    numberstyle=\scriptsize\color{light-gray}\textsf,
+    numberstyle=\footnotesize\color{light-gray}\textsf,
     moredelim=**[is][\color{blue}\bf\ttfamily]{`}{`},
@@
 \newtheorem{rem}[thm]{Remark} 
+%\renewcommand*{\proofname}{Proof Sketch}
 
@@
 
+\newcommand{\val}{{\sf val}}
+
+\newcommand{\rcheck}[4]{ {\sf rcheck}[#1](#2;#3;#4) }
+
 \newcommand{\strin}[1]{\sistr{#1}}
-\newcommand{\rsconcat}[2]{{\sf rconcat}(#1,#2)} \newcommand{\rconcat}[2]{{\sf rconcat}(#1,#2)} % lots of legact naming around these parts..
+\newcommand{\rsconcat}[2]{{\sf rconcat}(#1;#2)} \newcommand{\rconcat}[2]{{\sf rconcat}(#1;#2)} % lots of legact naming around these parts..
 
@@
 % target language terms.
-\newcommand{\tsubst}[3]{{\sf replace}(#1,#2,#3)} \newcommand{\metareplace}[3]{{\sf replace}(#1,#2,#3)} % TODO-nrf rename the commands. Lots of legacy naming around these parts...
-
-\newcommand{\tcheck}[2]{{\sf check}(#1, #2)}
+\newcommand{\tsubst}[3]{{\sf preplace}(#1;#2;#3)} \newcommand{\metareplace}[3]{{\sf replace}(#1;#2;#3)} % TODO-nrf rename the commands. Lots of legacy naming around these parts...
+
+\newcommand{\tcheck}[4]{{\sf pcheck}(#1; #2; #3; #4)}
 \renewcommand{\tstr}[1]{{{\sf str}[#1]}}
-\newcommand{\preplace}[3]{{\sf preplace}(#1,#2,#3)}
-\newcommand{\tconcat}[2]{{\sf concat}(#1,#2)} \newcommand{\concat}[2]{{\sf concat}(#1,#2)} % lots of legacy naming around these parts...
+\newcommand{\preplace}[3]{{\sf preplace}(#1;#2;#3)}
+\newcommand{\tconcat}[2]{{\sf pconcat}(#1;#2)} \newcommand{\concat}[2]{{\sf concat}(#1;#2)} % lots of legacy naming around these parts...
 
@@
 % Meta-theoretic functions
-\newcommand{\dosubst}[3]{{\tt replace}(#1,#2,#3)}
-\newcommand{\lsubst}[3]{{\tt lsubst}(#1,#2,#3)} % This used to render lreplace(...) so there're probably mistkes wherever this command was used now.
-\newcommand{\lreplace}[3]{{\tt lreplace}(#1, #2, #3)}
+\newcommand{\lsubst}[3]{{\sf replace}(#1;#2;#3)} % This used to renderlreplace(...) so there're probably mistkes wherever this command was used now.
+\newcommand{\lreplace}[3]{{\sf lreplace}(#1; #2; #3)}
+
+\newcommand{\sctx}{\Psi} % Context for external typing
+\newcommand{\tctx}{\Theta} % Context for internal typing
+\newcommand{\ereduces}{\Downarrow}
+
+\newcommand{\strcase}[3]{ {\sf rstrcase}(#1; #2; #3)}
+\newcommand{\pstrcase}[3]{ {\sf pstrcase}(#1; #2; #3)}
+
+\newcommand{\lhead}[1]{ {\sf lhead}(#1) }
+\newcommand{\ltail}[1]{ {\sf ltail}(#1) }
 
 % Judgements
-\newcommand{\err}{\ {\sf err}}
 \newcommand{\trden}[1]{\llbracket #1 \rrbracket} % = Translation Denotation.
@@
 % Names of type systems presented in paper
-\newcommand{\lcs}{\lambda_{CS}}
-
-\setlength{\grammarindent}{3em}
+\newcommand{\lcs}{\lambda_{S}}
+
+\setlength{\grammarindent}{2em}
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
 
+\usepackage[compact]{titlesec}
+\titlespacing*{\section}{0pt}{3pt}{0pt}
+\titlespacing*{\subsection}{0pt}{3pt}{0pt}
+\titlespacing{\subsubsection}{0pt}{3pt}{0pt}
+\titlespacing{\paragraph}{0pt}{3pt}{5pt}
+
 \begin{document}
@@
 
-\title{A Type System for String Sanitation\\Implemented Inside a Python}
-
-%\authorinfo{~}{~}{~}
-%dep_docclass\authorinfo{Nathan Fulton\and Cyrus Omar\and Jonathan Aldrich}
-%dep_docclass           {School of Computer Science\\
-%dep_docclass            Carnegie Mellon University}
-%dep_docclass           {\{nathanfu, comar, aldrich\}@cs.cmu.edu}
+\permissiontopublish
+\conferenceinfo{PSP~'14}{October 21, 2014, Portland, OR, USA.}
+\copyrightyear{2014}
+\copyrightdata{978-1-4503-2296-6/14/10}
+\doi{2687148.2687152} 
+\title{Statically Typed String Sanitation Inside a Python}
+%\numberofauthors{3}
+%\author{
+%  \alignauthor
+%  Nathan Fulton
+%  \alignauthor
+%  Cyrus Omar
+%  \alignauthor
+%  Jonathan Aldrich
+%  \and
+%  \affaddr{Carnegie Mellon University, Pittsburgh, PA}\\
+%  %\affaddr{Pittsburgh, PA}\\
+%  \email{\{nathanfu, comar, aldrich\}@cs.cmu.edu}
+%}
+%%\authorinfo{~}{~}{~}
+\authorinfo{Nathan Fulton\and Cyrus Omar\and Jonathan Aldrich}
+          {Carnegie Mellon University \\ 
+        \{nathanfu, comar, aldrich\}@cs.cmu.edu}
+%
 
@@
 \begin{abstract}
-%Evidence suggests that programmers are reluctant to adopt new languages to
-Security-oriented programming languages do not enjoy wide adoption.
-Conversely, library and framework-based approaches to security are pervasive.
-Unfortunately, libraries and frameworks are only effective solutions if they are implemented and used properly.
-In this paper, we propose that extensible programming languages provide a promising mechanism for increasing the amount of trust that can be place in libraries and frameworks.
-To demonstrate this approach, we present a conservative extension to the simply-typed lambda calculus for checking the correctness of input sanitation algorithms.
-We also demonstrate how this language's security guarantees are preserved under translation into an underlying language containing a regular expression library.
-We believe this approach -- complementing existing techniques with light-weight, composable type-based analyses -- constitutes a promising compromise between usability, potential for industry adoption, and theoretically grounded safety guarantees.
+Web applications must ultimately command systems like web browsers and database engines using strings.
+Strings derived from improperly sanitized user input  can as a result be a vector for command injection attacks.
+In this paper, we introduce \emph{regular string types}, which classify strings constrained statically to be in a regular language specified by a regular expression.
+Regular strings support standard string operations like concatenation and substitution, as well as safe coercions, so they  can be used to implement, in an essentially conventional manner, the pieces of a web application or framework that  handle strings arising from user input.
+Simple type annotations at function interfaces can be used to statically verify that sanitization has been performed correctly without introducing redundant run-time checks.
+We specify this type system first as a minimal typed lambda calculus, $\lambda_{RS}$.
+
+To be practical, adopting a specialized type system like this should not require the adoption of a new programming language.
+Instead, we advocate for extensible type systems: new type system fragments like this should be implemented as libraries atop a mechanism that guarantees that they can be safely composed.
+We support this with two contributions.
+First, we specify a translation from $\lambda_{RS}$ to a calculus with only standard strings and regular expressions.
+Then, taking Python as a language with these constructs, we implement the type system together with the translation as a library using \texttt{typy}, an extensible static type system for Python.
 \end{abstract}
 
-%\category{D.3.2}{Programming Languages}{Language Classifications}[Extensible Languages]
-%\category{D.3.4}{Programming Languages}{Processors}[Compilers]
-%\category{F.3.1}{Logics \& Meanings of Programs}{Specifying and Verifying and Reasoning about Programs}[Specification Techniques]
-%\keywords
-%type-level computation, typed compilation
+%\category{D.3.2}{Programming Languages}{Language Classifications}[Extensible languages]
+\category{F.3.3}{Logics \& Meanings of Programs}{Studies of Program Constructs}[Type structure]
+%\category{F.4.3}{Logics \& Meanings of Programs}{Formal Languages}
+\keywords
+type systems, regular expressions, input sanitation, string sanitation, extensible languages, web security
+
 \section{Introduction}\label{intro}
-Improper input sanitation is a leading cause of security vulnerabilities in web applications [OWASP].
-Command injection attacks exploit improper input sanitation by inserting malicious code into an otherwise benign command.
-Modern web frameworks, libraries, and database abstraction layers attempt to ensure proper sanitation of user input.
-When these methods are unavailable or insufficient, developers implement custom sanitation techniques.
-Im both cases, sanitation algorithms are implemented using the language's regular expression capabilities and usually \emph{replace} potentially unsafe strings with equivalent escaped strings.
-
-In this paper, we present a type system for implementing and statically checking input sanitation techniques.
+Command injection vulnerabilities are among the most common and severe security vulnerabilities in modern web applications \cite{OWASP}.
+They arise because web applications, at their boundaries, control external systems using commands represented as  strings.
+For example, web browsers are controlled using HTML and Javascript sent from a server as a string, and database engines execute SQL queries also sent as strings.
+When these commands include substrings derived from user input, care must be taken to ensure that the user cannot  subvert the intended command.
+For example, a  SQL query constructed using string concatenation exposes a SQL injection vulnerability:
+\begin{lstlisting}[numbers=none]
+'SELECT * FROM users WHERE name="' + name + '"'
+\end{lstlisting}\vspace{-3px}
+If a malicious user enters the name \lstinline{'"; DROP TABLE users --'}, the entire database could be erased.
+
+To avoid this problem, developers are cautioned to \emph{sanitize} user input.
+For example, in this case, the developer (or, more often, a framework) might define a function \verb|sanitize| that escapes double quotes and existing backslashes   with a backslash, which SQL treats securely.
+Alternatively, it might HTML-encode special characters, which would head off both SQL injection attacks and cross-site scripting attacks.
+\emph{Guaranteeing} that user input has already been sanitized  before it is used to construct a command is challenging, especially because  sanitization and command construction may occur in separately developed and maintained components.
+%is often performed in a different component of a program than where commands are ultimately constructed.% Note, for example, that \verb|sanitize| is not idempotent, so it should only be called once.
+
+We observe that many such sanitization techniques can be specified using \emph{regular languages} \cite{cinderella}.
+For example, for the command constructed above to be secure, \verb|name| must be a string in the language described by the regular expression \verb!([^"\\]|(\")|(\\))*!
+-- a sequence of characters other than quotation marks and backslashes; these can only appear escaped.
+This conventional syntax for regular expression patterns can be understood to desugar, in a standard way, to the syntax for regular expressions shown in Figure \ref{fig:regex}, where $r \cdot r$ is sequencing and $r + r$ is  disjunction.
+We will work with this ``core'' for simplicity in the remainder.
+
+In this paper, we present a static type system that tracks the regular language a string belongs to (we identify regular languages by the notation $\mathcal{L}\{r\}$).
+For example, the output of \verb|sanitize| will not simply be \verb|string|, but rather \verb|stringin[r]|, where \verb|r| is the regular expression above.
+By leveraging closure and decidability properties of regular languages, the type system will be able to closely track the language a string belongs to through uses of a number of operations, including \emph{replacement} of substrings matching a given pattern.
+This makes it possible to implement sanitization functions -- like the one just described -- in a conventional manner.
+The result is a system where the fact that a string has been {correctly} sanitized becomes manifest in its type.
+Missing calls to sanitization functions can thus be detected statically, and, crucially, so can \emph{incorrectly implemented sanitization functions} (i.e.
+these functions need not be trusted).
+These guarantees require run-time checks only when going from less precise to more precise types.
+%implementing and statically checking input sanitation techniques.
 Our solution suggests a more general approach to the integration of security concerns into programming language design.
@@
 
-To demonstrate this approach, we present a simply typed lambda calculus with \emph{constrained strings}; that is, a set of string types parameterized by regular expressions.
-If $s : \stringin{r}$, then $s$ is a string matching the language $r$.
-Additionally, we include an operation $\rreplace{r}{s_1}{s_2}$ which corresponds to the replace mechanism available in most regular expression libraries; that is, any substring of $s_1$ matching $r$ is replaced with $s_2$.
-The type of this expression is the computed, and is likely ``smaller" or more constrained than the type of $s_1$.
-Libraries, frameworks or functions which construct and execute commands containing input can specify a safe subset $\stringin{r_{\sf spec}}$ of strings, and input sanitation algorithms can construct such a string using ${\sf rreplace}$ or, optionally, by coercion (in which case a runtime check is inserted).
-We also show how this simple can be translated into a host language containing a regular expression library such that the safety guarantee of the extended language is preserved.
-
-Summarily, we present a simple type system extension which ensures the absence of input sanitation vulnerabilities by statically checking input sanitation algorithms which use an underlying regular expression library.
-This approach is \emph{composable} in the sense that it is a conservative extension.
-This approach is also \emph{complementary} to existing input sanitation techniques which use string replacement for input sanitation.
-
-%%
-% This is the paragraph which ties the paper back to the CFP.
-%
-% For refernce, here's a copy of the cfp:
-%The importance of security and privacy in software engineering is now a pressing concern. 
-%Over the last decade, many programmers have recognized the importance of including 
-%security and privacy requirements at project start. This has led to efforts 
-%in secure/security engineering and privacy engineering, which focus on guidelines and 
-%best practices that can be used at the design stage to create safer code. 
-%Unfortunately, these disciplines are not pervasive and still in their infancy.
-%The goal of this workshop is to make security and privacy first class citizens in programming and programming languages.
-
-%This forum aims to gather research and industry professionals to further the discussion on how to:
-%   codify the principles from secure and privacy engineering into programming language constructs and or tools, 
-%   create programming languages that have security and privacy as foundational tenets, and 
-%   create/codify constructs or tools that enables secure and privacy-preserving (business) operations.
-%   All submissions connected to the goals above, and their associated topics, are encouraged.
-%%
-
-\subsection{Related Work and Alternative Approaches}
-
-The input sanitation problem is well-understood.
-There exist a large number of techniques and technologies, proposed by both practitioners and researchers, for preventing injection-style attacks.
-In this section, we explain how our approach to the input sanitation problem differs from each of these approaches.
-More important than these differences, however, is our more general assertion that language extensibility is a promising approach toward consideration of security goals in programming lanugage design.
-
-Unlike \emph{frameworks and libraries} provided by languages such as Haskell and Ruby, our type system provides a \emph{static} guarantee that input is always properly sanitized before use.
-Doing so requires reasoning about the operations on regular languages corresponding to standard operations on strings; we are unaware of any production system which contains this form of reasoning.
-Therefore, even where frameworks and libraries provide a viable interface or wrapper around input sanitation, our approach is complementary because it ensurees the correctness of the framework or library itself.
-Furthermore, our approach is more general than database abstraction layers because our mechanism is applicable to all forms of command injection (e.g.
-shell injection or remote file inclusion).
-
-A number of research languages provide static guarantees that a program is free of input sanitation vulnerabilities [Jif][Ur/Web]. Our work differes from these contributions in the ways following:
-
+We will begin in Sec.
+\ref{calculus} by specifying this type system minimally, as a conservative extension of the simply typed lambda calculus called $\lambdas$.
+This allows us to specify the guarantees that the type system provides precisely.
+We also formally specify a translation from this calculus to a typed calculus with only standard strings and regular expressions, intending it as a guide to language implementors interested in building this feature into their own languages.
+This also demonstrates that no additional space overhead is required.
+
+Waiting for a language designer to build this feature in is unsatisfying in practice.
+%Moreover, we also face a ``chicken-and-egg problem'': justifying its inclusion into a commonly used language benefits from empirical case studies, but these are  difficult to conduct if developers have no way to use the abstraction in real-world code.
+We take the position that a better path forward for the community is to work within a programming language where such type system fragments can be introduced modularly and orthogonally, as libraries.
+
+In Sec.
+\ref{typy}, we show how to implement the type system and translation from Sec.
+\ref{calculus} using \texttt{typy}, an extensible static type system implemented as a library inside Python.
+\texttt{typy} leverages local type inference to control the semantics of literal forms, so regular string types can be introduced using string literals without any run-time overhead.
+Coercions that are known to be  safe due to a sublanguage relationship are performed implicitly, also without run-time overhead.
+This results in a \emph{usably secure} system: working with regular strings differs little from working with standard strings in a language that web developers have already widely adopted.
+
+We conclude after discussing related work in Sec. \ref{related}.
+
+\section{Regular String Types, Minimally}\label{calculus}
+\noindent
+%This will serve as the source language for our translation to a calculus with only standard strings and regular expressions, defined in Section~\ref{sec:tr}. 
+This section is organized as follows:
 \begin{itemize}
-  \item Our system is a light-weight solution to a single class of sanitation vulnerabilities (e.g. we do not address Cross-Site Scripting).
-  \item Our system is defined as a library in terms of an extensible type system, as opposed to a stand-alone language.
-Instead of introducing new technologies and methodologies for addressing security problems, we provide a light-weight static analysis which complements approaches developers already understand well.
-  \item Our implementation of the translation is implemented in Python and shares its grammar.
-Since Python is a popular programming language among web developers, the barrier between our research and adopted technologies is lower than for greenfield security-oriented languages.
+\item Sec. \ref{sec:rs} describes $\lambda_{RS}$ and its metatheory.% We also give proof outlines of  type safety and correctness theorems and relevant propositions about regular languages.
+\item Sec. \ref{sec:p} describes a simple target language, $\lambdap$, with a minimal regular expression library. In Section \ref{typy}, we will take Python to be such a language.
+\item Sec. \ref{sec:tr} describes the translation from $\lambdas$ to $\lambdap$ and ensures the correctness result from Sec. \ref{sec:rs} is preserved under this translation. 
 \end{itemize}
 
-We are also unaware of any extensible programming languages which emphasize applications to security concerns (TRUE?).
-
-Incorporating regular expressions into the type system is not novel.
-The XDuce system \cite{pierce} typechecks XML schemas using regular expressions.
-We differ from this and related work in at least two ways.
-First, our system  is defined within an extensible type system; second, and more importantly, we have demonstrated that regular expression types are applicable to the web security domain.
-
-In conclusion, our system is novel in at least two ways:
-\begin{itemize}
-\item The safety guarantees provided by libraries and frameworks in popular languages
-are not as (statically) justified as is often belived (or even claimed).
-\item Our extension is the first major demonstration of how an extensible type
-system may be used to provide light-weight, composable security analyses based upon
-idiomatic code.
-\end{itemize}
-
-\subsection{Outline}
-
-An outline of this paper follows:
-
-\begin{itemize}
-  \item In \S 2, we define the type system which is embedded in Ace. We include a type safety proof for the string
-    segment of this language and prove the correctness of a translation to an underlying langguage $P$. In our theory,
-    $P$ is a simply typed lambda calculus equipped with a minimal regular expression library; in an implementation, $P$
-    stands in for Python or another underlying general-purpose programming language.
-  \item In \S 3, we discuss our implemention of this translation as a type system extension  within the Ace programming language.
-\end{itemize}
-
-\section{A Type System for String Sanitation}
-
-The $\lambdas$ language is characterized by a type of strings indexed by regular
-expressions, together with operations on such strings which correspond to common
-input sanitation patterns.
-This section presents the grammar, typing rules and operational semantics for
-$\lambdas$ as well as an underlying language $\lambdap$.
-
-The system $\lambdas$ \ref{fig:lambdas} is the simply typed lambda calculus extended with \emph{regular expression types}, which are string
-types ensuring a string belongs to a specified language. For instance, 
-$S : \stringin{r}$ reads ``$s$ is a string matching $r$".
-the system includes an operation for replacing all instances of a pattern $r$ in a string $s_1$ with another string $s_2$. 
-Input sanitation algorithms -- as implemented by developers or within popular libraries and frameworks -- are often implemented in terms of this replace operation.
-For instance, a developer might all potentially unsafe charapcters with excaped versions of the same character.
-Regular expression types are used both to specify input sanitation algorithms, and at use sites as specifications.
-Note that runtime error states ($S \err$) are introduced by coercion, not by replacement.
-
-The language $\lambdap$ \ref{fig:lambdap} is a simple functional language extended with a minimal regular expression library.
-Any general purpose programming language could stand in for $\lambdap$; for instance, SML has a regular expression library.
-In an implementation, our correctness results are modulo the underlying language's correct implementation of regular expression matching (see P-E-Replace).
-
-Finally, we define a translation from our type system $\lambdas$ into $\lambdap$.
-
-%Note that we never insert an internal check where the type of the string implies
-%that a check must succeed. Furthermore, expensive calculations (such as language
-%inclusion or ${\sf rreplace}$) all occur at compile time.
-%Since sanitation problems are generally expressible with smalle, simple expressions,
-%we believe that the compile-time overhead is not significant enough to prohibiour target usecase. Informal experimentation with our implementation seems to
-%support this assertion.
-
-% Begin Grammar 
-% Removes the < > from grammar productions.
+The accompanying technical report \cite{psptr} contains more detailed proofs and further discussion of  design choices.% Figures \ref{fig:glambdas}-\ref{fig:tr} for the reader's convenience.
+
+\subsection{The Language of Regular Strings}\label{sec:rs}
+In this section, we define a typed lambda calculus with regular string types called $\lambda_{RS}$.
+Its syntax is specified in Figure \ref{fig:glambdas}, its static semantics in Figure \ref{fig:slambdas} and its evaluation semantics, given here in big-step style, in Figure \ref{fig:dlambdas}.
+
+There are two type constructors in $\lambdas$: $\rightarrow$ and $\textsf{stringin}$.
+Arrow types classify functions, which are introduced via lambda abstraction, $\lambda x.e$, and can be applied, written $e(e)$, in the usual way \cite{pfpl}.
+Regular string types are of the form $\stringin{r}$, where $r$ is a regular expression.
+Values of such regular string types take the form $\strin{s}$, where $s$ is a string (i.e.
+$s \in \Sigma^\star$, where the Kleene star is defined in the usual way).
+The rule \textsc{S-T-Stringin-I} requires that $s \in \lang{r}$.
+ %The remaining operations on terms of type $\stringin{r}$ preserve this property. %The remaining operators operate on regular strings.
+
+$\lambdas$ provides several familiar operations on strings. The type system  relates these operations over strings to corresponding operations over the  regular languages they belong to. 
+Since these operations over regular languages are known to be closed and decidable, we can use these operations as a basis for implementing and reasoning about sanitation protocols, as we will discuss below.
+
+%This section describes our approach, culminating in a Security Theorem.
+The main technical result of this paper shows that this Security Theorem is preserved when $\lambdas$ is translated to a more traditional language.
+Throughout this section, we refer to \textsc{Inference Rules} defined in the figures contained in this section.
+%\todo{remove tranditional, most and improve description}
+
+%In this section, we define $\lambdas$, describe each operation, and finally prove both type safety and a correctness result for the language.
+%The grammar for $\lambdas$ is defined in Figure \ref{fig:glambdas}. Typing rules \textsc{S-T-} are defined in Figure \ref{fig:slambdas} and a big-step semantics
+%is defined in Figure \ref{fig:dlambdas}
+
 \renewcommand{\grammarlabel}[2]{#1\hfill#2}
-
-\begin{figure}[h]
+\begin{figure}[t]
+\footnotesize
   \begin{grammar}
-<r> ::= $\epsilon$ | $.$ | $a$ | $r \cdot r$ | $r + r$ | $r*$ \hfill $a \in \Sigma$
+<$r$> ::= $\epsilon$ | $.$ | $a$ | $r \cdot r$ | $r + r$ | $r*$ \hfill $a \in \Sigma$
 
 \caption{Regular expressions over the alphabet $\Sigma$.}
+\label{fig:regex}
 \end{grammar}
 \end{figure}
-
-\begin{figure}[h]
+\begin{figure}[t]
+\footnotesize
   \begin{grammar}
-
-<$\psi$> ::=	...			\hfill	source types					\alt
-$\stringin{r}$
-
-<S> ::= ... \hfill source terms \alt
-      $\strin{s}$ \hfill $s \in \Sigma^{*}$ \alt
-      $\rsconcat{S}{S}$ \alt
-      $\sisubst{r}{S}{S}$ \alt
-      $\coerce{r}{S}$
-\caption{Syntax for the string sanitation fragment of our source language, $\lambdas$.}
+<$\sigma$> ::=  $\sigma \rightarrow \sigma$ | $\stringin{r}$    \hfill  source types
+\vspace{-8px}
+
+<$v$> ::= $\lambda x . e$ | $\strin{s}$ \hfill source values ($s \in \Sigma^{*}$) 
+\vspace{-8px}
+
+<$e$> ::= 
+      $v$ | $x$ | $e(e)$   \hfill source terms \alt 
+      $\rsconcat{e}{e}$ | $\strcase{e}{e}{x,y.e}$ \alt
+      $\rcoerce{r}{e}$ | $\rcheck{r}{e}{x.e}{e}$ | $\sisubst{r}{e}{e}$  
+      \caption{Syntax of $\lambda_{RS}$.}
+\label{fig:glambdas}
 \end{grammar}
@@
 
-\begin{figure}[h]
-\begin{grammar}
-
-<$\theta$> ::= ... \hfill target types \alt
-$\str$ \alt $\regex$
-
-<P> ::= ... \hfill target terms \alt
-  $\tstr{s}$ \alt
-  $\rx{r}$ \alt
-  $\tconcat{P}{P}$ \alt
-  $\preplace{P}{P}{P}$ \alt
-  $\tcheck{P}{P}$ 
-  \end{grammar}
-\caption{Syntax for the fragment of our target language, $\lambdap$, containing strings and statically constructed regular expressions.}
-\label{fig:lcsSyntax}
-\end{figure}
-
-%The $\lcs$ language gives static semantics for common regular expression library
-%functions. In this treatment, we include concatenation and filtering.
-%The ${\tt filter}$ function removes all instances of a regular expression in a string,
-%while concatenation ($+$) concatenates two strings.
-%
-%\subsection{Typing}
-%
-%The $\strin[r]$ type is parameterized by regular expressions; if $e:\strin[r]$,
-%then $e \in r$. Mapping from an arbitrary $\str$ to a $\strin[r]$ requires
-%defining an algorithm -- in terms of filter -- for converting a $\strin[.*]$
-%into a $\strin[r]$. The static semantics of the language defines the types of
-%operations on regular expressions in terms of well-understood properties about
-%regular lanugages; we recall these properties in section 3.
-%
-%\subsection{Dynamics}
-%
-%There are two evaluation judgements: $e:T \reduces e'$ and $e:T \treduces i$.
-%The $\reduces$ relation is between $\lcs$ expressions, while the $\treduces$
-%relation is a mapping from $\lcs$ expressions into internal language expressions
-%$i$ such that $i \ival$.
-%
-%Safety of the evaluation relation depends upon an injective mapping from $\lcs$ types info
-%internal language types. This relation, $h$, is defined below.
-%
-%\subsection{Type Safety}
-%
-%The type safety proof relies upon some assumptions about the type system and
-%dynamics of the internal language, as well as some properties of regular
-%languages.
-%
-%There must exist a translation from $\lcs$ types to the types of the internal
-%language. For the remainder of this paper, we call the type translation function
-%$h$.
-%
-%\begin{defn}[Type Translation Function $h$]
-%  The type translation function $h : Type \rightarrow IType$ is defined as follows:
-%  \begin{itemize}
-%    \item $\forall r. h(\strin[r]) = istr$
-%    \item $h(\str) = istr$
-%  \end{itemize}
-%\end{defn}
-%
-%Additionally, we assume that the internal language contains an implementation of strings,
-%together with operations for concatenation and filtering by regular expression.
-%
-%\begin{defn}[Types of internal values]
-%  Let $\tstrf{s}$ range over string literals and $r$ over regular expressions.
-%  Internal values are typed as follows:
-%\begin{itemize}
-%  \item If $e=\tstrf{s}$ then $e:istr$.
-%  \item If $e=\ifilter{r}{\tstrf{s}}$ then $e:istr$.
-%  \item If $e=\tstrf{s_1}+\tstrf{s_2}$ then $e:istr$.
-%\end{itemize}
-%\end{defn}
-%
-%For simplicity, we assume a fixed translation from
-%$\lcs$ regular expressions to regular expressions recognizable by the internal
-%language's regex library (in practice, a fixed translation is acceptable.)
-%To summarize, we assume an internal language containing 
-%a string type together with operations for string concatentation and filtering. We
-%expect closure over strings for both operations.
-%Finally, recall that ${\tt ifilter}$ is only needed for dynamic casts, which may be removed without
-%descreasing the expressivity or even usability of the language.
-%Finally, the semantics of the filter function are defined in terms of ${\tt rl\_filter}$,
-%which is a static version of ${\tt ifilter}$.
-%
-
-% I can't achieve the correct latout here. We should have something like:
-%       Lam_S Synax         Lam_P Syntax
-%       Lam_S Sematics      LAm_P Semantics
-%                Translation rules
-% Or, alternatively, we can flip reflect things so that the entire page
-% reads top-down (Lam_S syntax Lam_S smantics horizontally aligned w/ one
-% another).
-
-% Unfortunately, I can't get that to work. So, we're settling for something
-% slightly ugly, where the fitures appear in the incorrect order and the 
-% translation rules are on top.
-
-% The translations rules *just won't* go to the bottom of the page, so 
-% instead I am placing them at the top. Since the figures are out of order,
-% it's even more important that we get the number gith. therefore, the counter
-% is modified before and after this figure.
-\addtocounter{figure}{4}
-\begin{figure*}\label{fig:trans}
-
-$\fbox{\inferrule{}{\trden{S} = P}}$
+\subsubsection{Concatenation}
+
+The \textsc{S-T-Concat} rule is the simplest example of our approach.
+The rule is sound because the result of concatenating two strings, the first in $\lang{r_1}$ and the second in $\lang{r_2}$, will always be in the language $\lang{r_1\cdot r_2}$.
+The rule therefore relates string concatenation to sequential composition of regular expressions.
+
+\subsubsection{String Decomposition}
+Whereas concatenation allows the construction of large strings from smaller strings,  $\strcase{e}{e_0}{x,y.e_1}$  allows the decomposition, or elimination, of large strings into smaller strings.
+Intuitively, this operation branches based on whether a string is empty or not, exactly analagous to case analysis on lists in a functional language.
+The branch for a non-empty string ``peels off" the first character, binding it and the remainder of the string to specified variables.
+The evaluation rules \textsc{S-E-Case-$\epsilon$} and \textsc{S-E-Case-Concat} express this semantics.
+This construct can be used to implement a standard string indexing operation as a function, given a suitable definition of natural numbers (omitted).
+
+The typing rule \textsc{S-T-Case} must determine a suitable type for the head and tail of the string.
+The regular expression recognizing any one-character prefix of the strings in $\lang{r}$ is easily defined.
+
+\begin{figure}[t]
+\footnotesize
+$\fbox{\inferrule{}{\sctx \vdash e : \sigma}}$
+~~~~$\sctx ::= \emptyset \pipe \sctx, x : \sigma$
 \begin{mathpar}
-
-\inferrule[Tr-string]
-{ \ }
-{\trden{\strin{s}} = \tstr{s}}
-
-\inferrule[Tr-Concat]
-{ \trden{S_1} = P_1 \\ \trden{S_2} = P_2}
-{\trden{\rsconcat{S_1}{S_2}} = \tconcat{P_1}{P_2}}
-
-\inferrule[Tr-Subst]
-{ \trden{S_1} = P_1 \\ \trden{S_2} = P_2 }
-{\trden{ \sisubst{r}{S_1}{S_2} } = \tsubst{\rx{r}}{P_1}{P_2} }
-
-\inferrule[Tr-Coerce-Ok]
-{ S : \strin{r} \\ \lang{r'} \subseteq \lang{r}}
-{\trden{ \coerce{r'}{S} } = \tstr{s}}
-
-\inferrule[Tr-Coerce-NotOk]
-{\trden{S} = P \\ S : \strin{r} \\ \lang{r'} \not \subseteq \lang{r}}
-{\trden{\coerce{r'}{S}} = \tcheck{\rx{r'}}{P}}
-
-\end{mathpar}
-\caption{Translation from source terms (S) to target terms (P).
-The translation is type-directed in the Tr-Coerce cases.}
-\end{figure*}
-\addtocounter{figure}{-5}
-
-%
-% Begin typing/translation relation
-%
-
-\newcommand{\sctx}{\Psi} % Context for external typing
-\newcommand{\tctx}{\Theta} % Context for internal typing
-
-\newcommand{\ereduces}{\Downarrow}
-
-\begin{figure}[t]\label{fig:lambdas}
-\small
-$\fbox{\inferrule{}{\sctx \vdash S : \psi}}$
-~~~~$\sctx ::= \emptyset \pipe \sctx, x : \psi$
-\begin{mathpar}
+\inferrule[S-T-Var]
+{ x:\sigma \in \sctx }
+{ \sctx \vdash x:\sigma}
+
+\inferrule[S-T-Abs]
+{\sctx, x : \sigma_1 \vdash e : \sigma_2}
+{\sctx \vdash \lambda x.e : \sigma_1 \rightarrow \sigma_2}
+
+\inferrule[S-T-App]
+{\sctx \vdash e_1 : \sigma_2 \rightarrow \sigma \\ \sctx \vdash e_2 : \sigma_2}
+{\sctx \vdash e_1(e_2) : \sigma}
+
 \inferrule[S-T-Stringin-I]
@@
 \inferrule[S-T-Concat]
-{\sctx \vdash S_1 : \stringin{r_1} \\ \sctx \vdash S_2 : \stringin{r_2}}
-{\sctx \vdash \rsconcat{S_1}{S_2} : \stringin{r_1 \cdot r_2}}
-
-\inferrule[S-T-replace]
-{\sctx \vdash S_1 : \stringin{r_1} \\ \sctx \vdash S_2 : \stringin{r_2} \\\\ {\tt lsubst}(r,r_1,r_2) = r'}
-{\sctx \vdash \sisubst{r}{S_1}{S_2} : \stringin{r'}}
-
-\inferrule[S-T-coerce]
-{\sctx \vdash S : \stringin{r'}}
-{\sctx \vdash \coerce{r}{S} : \stringin{r}}
-
+{\sctx \vdash e_1 : \stringin{r_1} \\ \sctx \vdash e_2 : \stringin{r_2}}
+{\sctx \vdash \rsconcat{e_1}{e_2} : \stringin{r_1 \cdot r_2}}
+
+\inferrule[S-T-Case]
+{ \sctx \vdash e_1 : \stringin{r} \\
+  \sctx \vdash e_2 : \sigma \\
+  \sctx, x : \stringin{\lhead{r}}, y : \stringin{\ltail{r}} \vdash e_3 : \sigma
+}
+{
+  \sctx \vdash \strcase{e_1}{e_2}{x,y.e_3} : \sigma
+}
+
+\inferrule[S-T-SafeCoerce]
+{\sctx \vdash e : \stringin{r'} \\ \lang{r'} \subseteq \lang{r}}
+{\sctx \vdash \coerce{r}{e} : \stringin{r}}
+
+\inferrule[S-T-Check]
+{\sctx \vdash e_0 : \stringin{r_0} \\ \sctx, x:\stringin{r} \vdash e_1 : \sigma \\ \sctx \vdash e_2 : \sigma}
+{\sctx \vdash \rcheck{r}{e_0}{x.e_1}{e_2} : \sigma}
+
+\inferrule[S-T-Replace]
+{ \sctx \vdash e_1 : \stringin{r_1}} 
+{\sctx \vdash \sisubst{r}{e_1}{e_2} : \stringin{ \lreplace{r}{r_1}{r_2} }}
 \end{mathpar}
-
-\caption{Typing rules for our fragment of $\lambdas$. 
+\vspace{-10px}
+\caption{Typing rules for $\lambdas$. 
 The typing context $\sctx$ is standard.}
-\label{fig:etyping}
+\label{fig:slambdas}
 \end{figure}
 
-\begin{figure}[t]
-\small
-$\fbox{\inferrule{}{S \sreduces S}}$
-$\fbox{\inferrule{}{S \err}}$
+\begin{defn}[Definition of $\lhead{r}$] 
+{\footnotesize
+\begin{align*}
+\lhead{r}                 &= \lhead{r, \epsilon}
+\end{align*}}We use a two-argument auxiliary $\lhead{r, r'}$ in the definition of $\lhead{r}$ because when $r = q^*\cdot r'$,
+$\lhead{r}$ needs to ``remember" $r'$ in case $q$ is iterated zero times:
+{\footnotesize
+\begin{align*}
+\lhead{\epsilon, r'}      &= \epsilon \\                                        
+\lhead{a, r'}             &= a \\                                               
+\lhead{r_1\cdot r_2, r'}  &= \lhead{r_1, r_2} \\                                
+\lhead{r_1 + r_2, r'}     &= \lhead{r_1, r'} + \lhead{r_2, r'} \\               
+\lhead{r^*, r'}           &= \lhead{r', \epsilon} + \lhead{r, \epsilon}         
+\end{align*}} 
+\end{defn}
+
+Given this definition of $\lhead{r}$, regular expression derivatives \cite{bowzer} provide a useful tool for defining $\ltail{r}$.
+\begin{defn}[Brzozowski's Derivative]\label{def:derivative}
+  The \emph{derivative of $r$ with respect to $s$} 
+  is $\delta_s(r) = \{t | st \in \lang{r}\}$.
+\end{defn}
+
+Definition~\ref{def:derivative} is equivalent to the definition given in \cite{bowzer},
+although we refer the unfamiliar reader to \cite{owens}.
+Definition~\ref{def:derivative} is equivalent to Definition 3.1 in both papers.
+We now define $\ltail{r}$ using derivatives with respect to $\lhead{r}$.
+
+\begin{defn}[Definition of $\ltail{r}$]
+$\ltail{r}$ is defined in terms of $\lhead{r}$.
+Note that $\lhead{r} = a_1 + a_2 + ... + a_i$.
+We define $\ltail{r} = \delta_{a_1}(r) + \delta_{a_2}(r) + ... + \delta_{a_i}(r) + \epsilon$.
+\end{defn}
+
+The \textsc{S-T-Case} rule, which is defined in terms of these operations, thus relates the result of ``peeling off" the first character of a string to regular expression derivatives.
+
+\begin{figure}
+  \footnotesize
+$\fbox{\inferrule{}{e \sreduces v}}$
 \begin{mathpar}
+\inferrule[S-E-Abs]                                                             
+{ \ }                                                                           
+{\lambda x.e \sreduces \lambda x.e}
+
+\inferrule[S-E-App]
+{ e_1 \sreduces \lambda x . e_3 \\ e_2 \sreduces v_2 \\ [ v_2 / x ] e_3 \sreduces v }
+{ e_1(e_2) \sreduces v }
+
 \inferrule[S-E-RStr]
@@
 \inferrule[S-E-Concat]
-{S_1 \sreduces \strin{s_1} \\ S_2 \sreduces \strin{s_2}} 
-{\rsconcat{S_1}{S_2} \sreduces \strin{s_1 s_2}} % ???
+{e_1 \sreduces \strin{s_1} \\ e_2 \sreduces \strin{s_2}} 
+{\rsconcat{e_1}{e_2} \sreduces \strin{s_1 s_2}} % ???
+
+\inferrule[S-E-Case-$\epsilon$]
+{
+  e_1 \sreduces \strin{\epsilon} \\
+  e_2 \sreduces v_2
+}
+{
+  \strcase{e_1}{e_2}{x,y.e_3} \sreduces v_2
+}
+
+\inferrule[S-E-Case-Concat]
+{
+  e_1 \sreduces \strin{a s} \\
+  [\rstr{a},\rstr{s} / x,y]e_3 \sreduces v_3
+}
+{
+  \strcase{e_1}{e_2}{x,y.e_3} \sreduces v_3
+}
+
+\inferrule[S-E-SafeCoerce]
+{e \sreduces \strin{s}}
+{\coerce{r}{e} \sreduces \strin{s}}
+
+\inferrule[S-E-Check-Ok]
+{e \sreduces \strin{s} \\ s \in \lang{r} \\ [\strin{s} / x]e_1 \sreduces v}
+{
+  \rcheck{r}{e}{x.e_1}{e_2} \sreduces v
+}
+
+\inferrule[S-E-Check-NotOk]
+{
+  e \sreduces \strin{s} \\ s \not \in \lang{r} \\
+  e_2 \sreduces v
+}
+{
+  \rcheck{r}{e}{x.e_1}{e_2} \sreduces v
+}
 
 \inferrule[S-E-Replace]
-{S_1 \sreduces \strin{s_1} \\ S_2 \sreduces \strin{s_2} \\ \dosubst{r}{s_1}{s_2} = s} 
-{\sisubst{r}{S_1}{S_2} \sreduces \sistr{s}}
-
-\inferrule[S-E-Coerce-OK]
-{S \sreduces \strin{s} \\ s \in \lang{r}}
-{\coerce{r}{S} \sreduces \strin{s}}
-
-\inferrule[S-E-Coerce-Err]
-{S \sreduces \strin{s} \\ s \not \in \lang{r}}
-{\coerce{r}{S} \err}
-
+{e_1 \sreduces \strin{s_1} \\ e_2 \sreduces \strin{s_2}} 
+{\sisubst{r}{e_1}{e_2} \sreduces \sistr{ \lsubst{r}{s_1}{s_2} }}
 \end{mathpar}
-\caption{Big step semantics for our fragment of $\lambdas$.
-Error propagation rules are omitted.}
-\label{fig:eeval}
+\vspace{-10px}
+\caption{Big step semantics for $\lambdas$}.
+\label{fig:dlambdas}
 \end{figure}
 
-\newpage
+\subsubsection{Coercion}
+
+The $\lambdas$ language supports two forms of coercion. Safe coercions, written $\rcoerce{r}{e}$, allow passage to a ``smaller'' regular language. Such coercions will always succeed. 
+% This form of coercion is useful because the replacement operation (introduced below in Section~\ref{sec:replace}) is often more conservative than absolutely necessary.
+Conversely, checked coercions, $\rcheck{r}{e_0}{x.e_1}{e_2}$, 
+allow for passage to regular languages that are not necessarily smaller. Checked coercions branch based on whether the coercion succeeded or not.
+
+The rule 
+ \textsc{S-T-SafeCoerce} checks for language inclusion, which we write $\lang{r_1} \subseteq \lang{r_2}$.
+Language inclusion is decidable.
+As a result, the rule \textsc{S-E-SafeCoerce} does not need to perform any dynamic checks.
+The rule
+\textsc{S-T-Check}, conversely, does not perform any static checks on the two languages involved, only checking that the two branches have the same type.
+The checks are performed dynamically, by rules \textsc{S-E-Check-Ok} and \textsc{S-E-Check-NotOk}.
+
+In our calculus, both forms of coercion are explicit.
+For safe coercions, it is often useful for the coercion to be performed implicitly.
+This can be seen as a form of subtyping for regular string types \cite{gunter, fulton13, xhaskell2}.
+In practice, subtyping will be crucial to the usability of our system due to the nature of the replacement operator.
+For simplicity, we do not specify subtyping in our core calculus; however, subtyping for regular strings is present in previous treatments of our work \cite{fulton13}.
+The implementation discussed in Section \ref{typy} also provides subtyping between regular string types.
+
+\subsubsection{Replacement}\label{sec:replace}
+The premier operation for working with regular strings in $\lambdas$ is  replacement, written $\sisubst{r}{e_1}{e_2}$. 
+It behaves analogously to \lstinline{str_replace} in PHP and \lstinline{String.replace} in Java, differing in that the replacement pattern $r$ must be statically given.
+The evaluation rule \textsc{S-E-replace} is defined in terms of the metafunction $\lsubst{r}{s_1}{s_2}$.
+
+\begin{defn}[$\sf{replace}$]
+$\lsubst{r}{s_1}{s_2} = s$ such that all substrings of $s_1$ in $\lang{r}$ are replaced with $s_2$.
+\end{defn}
+The typing rule \textsc{S-T-Replace} thus requires computing the regular language of the string resulting from this operation given knowledge of the languages of $s_1$ and $s_2$, written as $\lreplace{r}{r_1}{r_2}={r'}$.
+
+Given an automata-oriented interpretation of regular languages, it may be helpful to think in terms of replacing sub-automata.
+A complete definition of ${\sf lreplace}$ would consist of a rewrite system based on this intuition, with correctness and termination proofs, which is beyond the scope of this paper.
+Instead, we provide an abstract definition of the operation and state necessary properties.
+
+\begin{defn}[$\sf{lreplace}$]
+  $\lreplace{r}{r_1}{r_2}$ relates $r, r_1,$ and $r_2$ to
+  a language $r'$ containing all strings of $r_1$ except that any substring $s_{pre} s s_{post} \in \lang{r_1}$ where $s \in \lang{r}$
+  is replaced by the set of strings $s_{pre} s_2 s_{post}$ for all $s_2 \in \lang{r_2}$ (the prefix and postfix positions may be empty). This procedure saturates the string.
+\end{defn}
+
+% \begin{prop}[Closure] \label{thm:total}
+%   If $\lang{r}, \lang{r_1}$ and $\lang{r_2}$ are regular languages, then $\lang{\lreplace{r}{r_1}{r_2}}$ is also a regular language.
+% \end{prop}
+% \begin{prf}
+% Algorithms for the inclusion problem may be adopted to identify any sublanguage $x \subseteq r$ of $r_1$.
+% The language $x_{pre}$ can be computed by taking total derivatives until the remaining language
+% equals $x$; $x_{post}$ can be computed in a similar way after reversal.
+% Then $r'$ is $x_{pre}r_2x_{post}$.
+% \end{prf}
+
+\begin{prop}[Replacement Correspondence] \label{thm:substcorrespondence}
+  Given strings $s_1 \in \lang{r_1}$ and $s_2 \in \lang{r_2}$ we have $$\lsubst{r}{s_1}{s_2} \in \lang{\lreplace{r}{r_1}{r_2}}$$
+\end{prop}
+
+\subsubsection{Metatheory of $\lambdas$}
+
+In this section, we establish some basic metatheoretic properties of $\lambdas$.
+These rely upon the definitions and propositions given above and some basic properties of regular languages.
+
+\begin{lem}[Properties of Regular Languages.] \label{thm:regexprops}
+~
+\begin{enumerate}
+\item 
+If $s_1 \in \lang{r_1}$ and $s_2 \in \lang{r_2}$ then $s_1s_2 \in \lang{r_1\cdot r_2}$.
+\item 
+For all strings $s$ and regular expressions $r$, either $s \in \lang{r}$ or $s \not \in \lang{r}$.
+%\item 
+%Regular languages are closed under reversal.
+\end{enumerate}
+\end{lem}
+
+If any of these properties are unfamiliar, the reader may refer to a standard text on the subject \cite{cinderella}.
+
+% We also need to establish a well-formedness lemma, which simply expresses the totality of the operations over regular expressions used as premises in the rules in Figure \ref{fig:slambdas}. 
+% \begin{lem}
+%   If $\sctx \vdash e : \stringin{r}$ then $r$ is a well-formed regular expression.
+% \end{lem}
+
+Type preservation for $\lambdas$ requires validating that the statics are consistent with the dynamics.
+We give a full proof of type safety for a variant of the calculus with a small step semantics in \cite{psptr}, but for concision, it is more straightforward to explain the semantics with a big step semantics here.
+
+% We avoid the most significant issues inherent to proofs related to big-step semantics by avoiding non-termination in our specification.
+% The simply typed lambda calculus terminates, and our conservative extension clearly
+% terminates because new binding structures are not introduced and termination (i.e.
+decidability) of ${\sf subst}$ and ${\sf lreplace}$ is known.
+Even in a non-terminating calculus, type safety would hold, but a more careful treatment or treatment for a language with non-termination would need to proceed by either a coinductive argument \cite{coinductionoo} or  with a small-step semantics.
+
+\begin{thm}[Type Preservation.] \label{thm:typesafety}
+  If $\emptyset \vdash e : \sigma$ 
+  and $e \sreduces v$ then $\emptyset \vdash v : \sigma$.
+\end{thm}
+\begin{prf}
+By induction on the typing relation.
+The \textsc{S-T-Concat} case requires Lemma \ref{thm:regexprops}.1, the \textsc{S-T-Check} case requires Lemma \ref{thm:regexprops}.2 
+and the S-T-Replace case appeals to Proposition \ref{thm:substcorrespondence}.
+\end{prf}
+
+We can also define a canonical forms lemma for regular strings.
+
+\begin{lem}[Canonical Forms for Regular Strings]\label{thm:cfs}
+  If $\emptyset \vdash v : \stringin{r}$ then $v=\rstr{s}$ and $s \in \lang{r}$. 
+\end{lem}
+\begin{prf}
+The only typing rule that applies is \textsc{S-T-Stringin-I}. The conclusion is the first premise.
+\end{prf}
+\subsubsection{The Security Theorem}\label{sec:securitythm}
+
+The chief benefit of $\lambdas$ is its security theorem, which states that any term of
+a regular string type that evaluates to a value will evaluate to a regular string recognized by the regular language corresponding to the regular expression the type is indexed by.
+This is beneficial because this ensure that membership in a regular language known to be secure becomes manifest in the type of the string, rather than being a property that must be established extralinguistically.
+
+\begin{thm}[Correctness of Input Sanitation for $\lambdas$]\label{thm:scorrect}
+  If  $\emptyset \vdash e : \stringin{r}$ and $e \sreduces \rstr{s}$ then $s \in \lang{r}$.
+\end{thm}
+\begin{prf}
+  The theorem follows directly from type preservation and canonical forms above.
+\end{prf}
+
+\subsection{Target Language}\label{sec:p}
+\begin{figure}[t]
+\footnotesize
+  \begin{grammar}
+
+<$\tau$> ::= $\tau \rightarrow \tau$ | $\str$ | $\regex$ \hfill target types
+\vspace{-8px}
+
+<$\dot{v}$> ::= $\lambda x . \iota$ | $\tstr{s}$ | $\rx{r}$ \hfill target values ($s \in \Sigma^{*}$)
+\vspace{-8px}
+
+<$\iota$> ::= $\dot{v}$ | $x$ | $\iota(\iota)$ \hfill target terms \alt
+   $\tconcat{\iota}{\iota}$ | $\pstrcase{\iota}{\iota}{x,y.\iota}$ \alt
+   $\tcheck{\iota}{\iota}{\iota}{\iota}$ | $\preplace{\iota}{\iota}{\iota}$
+\end{grammar}
+\vspace{-5px}
+\caption{Syntax of the target language, $\lambdap$, containing strings and statically constructed regular expressions.}
+\label{fig:lcsSyntax}
+\end{figure}
+Our next major technical result, stated in Sec. \ref{sec:tr}, establishes that the security property
+is preserved under translation into $\lambdap$. The system $\lambdap$ is another straight-forward extension of a simply typed lambda calculus
+with a $\str$ type and a $\regex$ type, as well as 
+some operations -- such as concatenation and replacement -- found in the standard libraries of many programming languages.
+The operations of $\lambdap $ correspond to ``run-time'' versions of the operations performed statically in $\lambdas$ in a way made precise by the translation rules described in the next section. 
+%Unlike $\lambdas$, $\lambdap$ does not statically track the effects of string operations.
+The language $\lambdap$ is so-called because it is reminscent of popular web programming languages,
+such as \textbf{P}ython or \textbf{P}HP, albeit statically typed. We will discuss an implementation within the former in the next section.
+
+The grammar of $\lambdap$ is defined in Figure~\ref{fig:lcsSyntax}.
+The typing rules \textsc{P-T-} are defined in Figure~\ref{fig:slambdap}
+and a big-step semantics is defined by the rules \textsc{P-E-} in Figure~\ref{fig:dlambdap}.
+The semantics are straightforward given the discussion of the corresponding operations in the previous section.
+
+\renewcommand{\grammarlabel}[2]{#1\hfill#2}
+
+\subsubsection{Safety}
 
 \begin{figure}[t]\label{fig:lambdap}
-\small
-$\fbox{\inferrule{}{\tctx \vdash P : \theta}}$
-~~~~$\tctx ::= \emptyset \pipe \tctx, x : \theta$
-
+\footnotesize
+$\fbox{\inferrule{}{\tctx \vdash \iota : \tau}}$
+~~~~$\tctx ::= \emptyset \pipe \tctx, x : \tau$
 \begin{mathpar} 
+\inferrule[P-T-Var]
+{ x:\tau \in \tctx }
+{ \tctx \vdash x:\tau }
+
+\inferrule[P-T-Abs]
+{\tctx, x : \tau_1 \vdash \iota_2 : \tau_2}
+{\tctx \vdash \lambda x.\iota_2 : \tau_1 \rightarrow \tau_2}
+
+\inferrule[P-T-App]
+{\tctx \vdash \iota_1 : \tau_2 \rightarrow \tau \\ \tctx \vdash \iota_2 : \tau_2}
+{\tctx \vdash \iota_1(\iota_2) : \iota}
+
 \inferrule[P-T-String]
@@
 \inferrule[P-T-Concat]
-{\tctx \vdash P_1 : \str \\ \tctx \vdash P_2 : \str}
-{\tctx \vdash \tconcat{P_1}{P_2} : \str}
+{\tctx \vdash \iota_1 : \str \\ \tctx \vdash \iota_2 : \str}
+{\tctx \vdash \tconcat{\iota_1}{\iota_2} : \str}
+
+\inferrule[P-T-Case]
+{\tctx \vdash \iota_1 : \str \\ \tctx \vdash \iota_2 : \tau \\ \tctx, x:\str, y:\str \vdash \iota_3:\tau}
+{\tctx \vdash \pstrcase{\iota_1}{\iota_2}{x,y.\iota_3} : \tau}
 
 \inferrule[P-T-Replace]
-{\tctx \vdash P_1 : \regex \\ \tctx \vdash P_2 : \str \\ \tctx \vdash P_3 : \str }
-{\tctx \vdash \preplace{P_1}{P_2}{P_3} : \str}
+{\tctx \vdash \iota_1 : \regex \\ \tctx \vdash \iota_2 : \str \\ \tctx \vdash \iota_3 : \str }
+{\tctx \vdash \preplace{\iota_1}{\iota_2}{\iota_3} : \str}
 
 \inferrule[P-T-Check]
-{\tctx \vdash P_1 : \regex \\ \tctx \vdash P_2 : \str \\ }
-{\tctx \vdash \tcheck{P_1}{P_2} : \str}
+{\tctx \vdash \iota_r : \regex \\ \tctx \vdash \iota_1 : \str \\ \tctx \vdash \iota_2 : \sigma \\ \tctx \vdash \iota_3 : \sigma}
+{\tctx \vdash \tcheck{\iota_r}{\iota_1}{\iota_2}{\iota_3} : \sigma}
 \end{mathpar}
-\caption{Typing rules for our fragment of $\lambdap$.
+\vspace{-10pt}
+\caption{Typing rules for $\lambdap$.
 The typing context $\tctx$ is standard.}
-\label{fig:etyping}
+\label{fig:slambdap}
 \end{figure}
+Type preservation for $\lambdap$ is essentially trivial, but is necessary in order to establish the 
+correctness of our translation.
+Again, we give a small step semantics and address type safety more completely in \cite{psptr}.
+
+\begin{thm}[Safety for $\lambdap$] If $\emptyset \vdash \iota : \tau$ 
+  and $\iota \sreduces \dot v$ then $\emptyset \vdash \dot v : \tau$.
+\end{thm}
+
+We can also define canonical forms for regular expressions and strings in the usual way:
+
+\begin{lem}[Canonical Forms for Target Language]
+If $\emptyset \vdash \dot{v} : \tau$ then 
+\begin{enumerate}
+\item 
+If $\tau=\regex$ then $\dot{v}= \rx{r}$ such that $r$ is a well-formed regular expression. 
+\item 
+If $\tau=\str$ then $\dot{v} = \tstr{s}$.
+\end{enumerate}
+\end{lem}
+
+\subsection{Translation from $\lambdas$ to $\lambdap$}\label{sec:tr}
+
+The translation from $\lambdas$ to $\lambdap$ is defined in Figure \ref{fig:tr}.
+The coercion cases are most interesting. If the safety of coercion in manifest in the
+types of the expressions, then no runtime check is inserted (\textsc{Tr-SafeCoerce}).
+If the safety of coercion is not manifest in the type, then a check is inserted (\textsc{Tr-Check}). Note that regular strings translate to strings directly; there is no space overhead.
+
+The translation correctness theorem guarantees that the translation is type preserving and that semantics of the original code and translation coincide, following the treatment of compilation pioneered by the TIL compiler for SML \cite{tarditi+:til-OLD}.
+Note that we apply the translation inline in judgements for concision when convenient.
+
+\begin{thm}[Translation Correctness]\label{thm:trcorrect}
+  If $\tctx \vdash e : \sigma$ then 
+  there exists an $\iota$ such that $\trden{e} = \iota$
+  and $\trden{\tctx} \vdash \iota : \trden{\sigma}$.
+  Furthermore, if $e \sreduces v$ then
+  $\iota \treduces \dot{v}$ such that
+  $\trden{v} = \dot{v}$.
+\end{thm}
+\begin{prf}
+The proof proceeds by induction on the typing relation for $e$.
+We choose an $\iota$ based on the syntactic form in $\lambdap$ corresponding to the form under consideration (e.g.
+we  choose {\sf replace} when considering {\sf sreplace}).
+The proof proceeds by our type safety theorems and an appeal to the induction hypothesis.
+\end{prf}
 
 \begin{figure}[t]
-\small
-$\fbox{\inferrule{}{P \treduces P}}$
-$\fbox{\inferrule{}{P \err}}$
-
+\footnotesize
+$\fbox{\inferrule{}{\iota \treduces \dot{v}}}$\vspace{-20px}
 \begin{mathpar}
+\hspace{30px}\inferrule[P-E-Abs]
+{ \ }
+{\lambda x.\iota \sreduces \lambda x.\iota}
+~~~~~~
+\inferrule[P-E-App]
+{ \iota_1 \sreduces \lambda x . \iota_3 \\  \iota_2 \sreduces \dot{v}_2 \\ [\dot{v}_2 / x] \iota_3 \sreduces \dot{v}_3}
+{ \iota_1(\iota_2) \sreduces \dot{v}_3}
+
 \inferrule[P-E-Str]
@@
 \inferrule[P-E-Concat]
-{P_1 \treduces \tstr{s_1} \\ P_2 \treduces \tstr{s_2}} 
-{\tconcat{P_1}{P_2} \treduces \tstr{s_1 s_2}} % ???
+{\iota_1 \treduces \tstr{s_1} \\ \iota_2 \treduces \tstr{s_2}} 
+{\tconcat{\iota_1}{\iota_2} \treduces \tstr{s_1 s_2}} % ???
+
+\inferrule[P-E-Case-$\epsilon$]
+{
+  \iota_1 \treduces \tstr{ \epsilon } \\
+  \iota_2 \treduces \dot{v_2} \\
+}
+{
+  \pstrcase{\iota_1}{\iota_2}{x,y.\iota_3} \treduces \dot{v_2}
+}
+
+\inferrule[P-E-Case-Concat]
+{
+  \iota_1 \treduces \tstr{as} \\
+  [\tstr{a},\tstr{s}/x,y] \iota_3 \treduces \dot{v} 
+}
+{
+  \pstrcase{\iota_1}{\iota_2}{x,y.\iota_3} \treduces \dot{v}
+}
 
 \inferrule[P-E-Replace]
-{P_1 \treduces \rx{r} \\ P_2 \treduces \tstr{s_2} \\ P_3 \treduces \tstr{s_3} \\ \dosubst{r}{s_2}{s_3} = s} 
-{\preplace{P_1}{P_2}{P_3} \treduces \tstr{s}}
+{\iota_1 \treduces \rx{r} \\ \iota_2 \treduces \tstr{s_2} \\ \iota_3 \treduces \tstr{s_3} \\ \lsubst{r}{s_2}{s_3} = s} 
+{\preplace{\iota_1}{\iota_2}{\iota_3} \treduces \tstr{s}}
 
 \inferrule[P-E-Check-OK]
-{P_1 \treduces \rx{r} \\ P_2 \treduces \strin{s} \\ s \in \lang{r}}
-{\tcheck{P_1}{P_2} \treduces \tstr{s}}
-
-\inferrule[P-E-Check-Err]
-{P_1 \treduces \rx{r} \\ P_2 \treduces \tstr{s} \\ s \not \in \lang{r}}
-{\tcheck{P_1}{P_2} \err}
+{
+  \iota_r \treduces \rx{r} \\
+  \iota \treduces \tstr{s} \\ 
+s \in \lang{r} \\ 
+\iota_1 \treduces \dot{v_1} 
+}
+{
+  \tcheck{\iota_r}{\iota}{\iota_1}{\iota_2} \treduces \dot{v_1}
+}
+
+\inferrule[P-E-Check-NotOK]
+{\iota_r \treduces \rx{r} \\ \iota \treduces \tstr{s} \\ s \not \in \lang{r} \\ \iota_2 \treduces \dot{v_2}}
+{\tcheck{\iota_r}{\iota}{\iota_1}{\iota_2} \treduces \dot{v_2}}
 \end{mathpar}
-\caption{Big step semantics for our fragment of $\lambdap$.
-Error propagation rules are omitted.}
-\label{fig:ieval}
+\vspace{-10px}
+\caption{Big step semantics for $\lambdap$}.
+\label{fig:dlambdap}
 \end{figure}
 
-\newpage
-\ 
-\newpage
-\ 
-\newpage
-
-\subsection{Properties of Regular Languages}
-
-Our type safety proofs for languages S and P and our translation correctness
-result all depend on some properties of regular languages. The crucial property
-is a relationship between string substitution -- which is available in any regular
-expression library -- and regular language substitution, which is a corresponding operation
-on languages instead of strings \ref{thm:substcorrespondence}.
-The decidability of language substitution is what enables static analysis of 
-sanitation algorithms implemented in terms of string replacement
-
-Throughout this section, we fix an alphabet $\Sigma$ over which strings $s$ and
-regular expressions $r$ are defined. throughout the paper, $\lang{r}$ refers to the
-language recognized by the expression $r$. This distinction between the expression
-and its language -- typically elided in the literature -- makes our definition and
-proofs about systems S and P more readable.
-
-\begin{lem}{Properties of Regular Languages and Expressions.} \label{thm:regexprops}
-The following are well-known properties of regular expressions which are necessary for our proofs:
-\begin{description}
-\item[(1):] If $s_1 \in \lang{r_1}$ and $s_2 \in \lang{r_2}$ then $s_1s_2 \in \lang{r_1r_2}$
-\item[(2):] For all strings $s$ and expressions $r$, either $s \in \lang{r}$ or $s \not \in \lang{r}$.
-\item[(3):] Regular languages are closed under complements and concatenation.
-\item[(4):] The regular expressions correspond bijectively to the regular languages.
-\end{description}
-\end{lem}
-
-\begin{defn}[$\tt{lsubst}$]
-The function $\lsubst{r}{s_1}{s_2}$ produces a string in which all substrings of $s_1$ matching $r$ are replaced with $s_2$.
-\end{defn}
-
-\begin{defn}[$\tt{lreplace}$]
-The function $\lreplace{r}{r_1}{r_2}$ produces a regular expression in which any sublanguage $\lang{r_1'}$ of $\lang{r_1}$ satisfying the condition $\lang{r_1'} \subseteq{\lang{r}}$ is replaced with $\lang{r_2}$.
-\end{defn}
-
-\begin{lem}{Closure and Totality of Replacement.} \label{thm:total}
-If $r, r_1$ and $r_2$ are regular expressions, then $\lreplace{r}{r_1}{r_2}$ is also a regular expression.
-\end{lem}
-\begin{proof}
-By induction on $r$ and closure properties of regular expressions.
-\end{proof}
-
-\begin{lem}{Substitution Correspondence.} \label{thm:substcorrespondence}
-If $s_1 \in \lang{r_1}$ and $s_2 \in \lang{r_2}$ then $\lsubst{r}{s_1}{s_2} \in \lreplace{r}{r_1}{r_2}$.
-\end{lem}
-\begin{proof}
-The proof proceeds by structural induction on $r$.
-
-\begin{description}
-\item[case $r=\alpha$.]
-If $s_1 = \alpha$ then $\alpha \in \lang{r_1}$ by assumption.
-Therefore, $\lsubst{r}{s_1}{s_2} = \lsubst{\alpha}{\alpha}{s_2} = s_2$ and
-$\lreplace{r}{r_1}{r_2} = \lreplace{\alpha}{r_1}{r_2}$. 
-Since $s_1 = \alpha$ and $\alpha \in \lang{r_1}$, $r_1 \iso \alpha | r_1'$ for some $r_1'$.
-Therefore, $\lreplace{\alpha}{r_1}{r_2} \iso \lreplace{\alpha}{\alpha|r_1'}{r_2}$ by Lemma X(1).
-Finally, $s_2 \in \lang{r_2}$ which implies $s_1 \in \lang{r_2|r'}$.
-If $s_1 \not = \alpha$ the $\lsubst{r}{s_1}{s_2} = s_1$ and $\lreplace{\alpha}{r_1}{r_2} = r_1$.
-
-\item[case $r = a|b$].
-Note that $\subst{a|b}{s_1}{s_2} = \lsubst{a}{\lsubst{b}{s_1}{s_2}}{s_2}$
-and $\lsubst{b}{s_1}{s_2} \in \lreplace{b}{r_1}{r_2}$ by induction.
-Therefore, $\lsubst{a|b}{s_1}{s_2}$ $\in \lreplace{a}{\lreplace{b}{r_1}{r_2}}{r_2}$ by induction and the definition of lreplace.
-Finally, applying definitions once more, $\lsubst{a|b}{s_1}{s_2} \in \lreplace{a|b}{r_1}{r_2}$.
-
-\item[case $r = ab$].
-By a similar argument to the disjunctive case.
-
-\item[case $r = a*$].
-By considering the once unwinding of $a*$, noting that $s_1$ and $s_2$ are finite.
-\end{description}
-\end{proof}
-
-% NRF The coercion lemma is no longer necessary, because it is a property of
-% "strip"-style sanitation rather than the more general case of replacement 
-% treated in this paper. However, we should mention this difference between my
-% undergraduate thesis and this paper in the related work section if we are
-% going tin clude a discussio of unpublished or quasi-published (e.g. src)
-% results.
-%\begin{lem}{Coercion Lemma.} \label{thm:coercionalemma}
-%Suppose that $R$ and $L$ are regular expressions, and that $s \in R$ is a finite string.  
-%Let $s' = \coerce(R,L,s)$ with all maximal substrings recognized by $L$ replaced with $\epsilon$.
-Then $s'$ is recognized by $(R \backslash L) + \emptyset$ and the construction of $R \backslash L$ is decidable.
-%\end{lem}
-%\begin{proof}
-%Let $F,G$ be FAs corresponding to $R$ and $L$, and let $G'$ be $G$ with its final states inverted (so that $G'$ is the complement of $L$).
-Define an FA $H$ as a DFA corresponding to the NFA found by combining $F$ and $G'$ such that $H$ accepts only if $R$ and $L'$ accept or if $s$ is empty (this construction may result in an exponential blowup in state size.)  Clearly, $H$ corresponds to $R \backslash L + \emptyset$.
-Thus, the construction of $R \backslash L + \emptyset$ is decidable.
-%
-%If $R \subset L$, $s' = \emptyset$.
-If $L \subset R$, either $s' = \emptyset$, or $s' \in R$ and $s' \not \in L$.
-If $R$ and $L$ are not subsets of one another, then it may be the case that $L$ recognizes part of $R$.
-Consider $L$ as the union of two languages, one which is a subset of $R$ and one which is disjoint.
-The subset language is considered above and the disjoint language is inconsequential.
-%\end{proof}
-
-% Why is this necessary?
-%\begin{lem}
-%  For all $r$, if $s_1 \in r_1$ and $s_2 \in r_2$ then $\subst{r}{s_1}{s_2} \subseteq \lsubst{r}{r_1}{r_2}$.
-%\end{lem}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-% BEGIN PROOFS ABOUT S AND P
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-\subsection{Safety of the Source and Target Languages}
-
-\begin{lem}
-  If $\sctx \vdash S : \stringin{r}$ then $r$ is a well-formed regular expression. 
-\end{lem}
-\begin{proof}
-  The only non-trivial case is S-T-Replace, which follows from \ref{thm:total}. 
-\end{proof}
-
-\begin{lem}
-  If $\tctx \vdash P : \regex$ then $P \treduces \rx{r}$ such that $r$ is a well-formed regular expression. 
-\end{lem}
-
-We now prove safety for the string fragment of the source and target languages.
-
-\begin{thm}{Safety for the String Fragment of P.}
-  Let $S$ be a term in the source language. If $\sctx \vdash S : \stringin{r}$ 
-  then $S \sreduces \rstr{s}$ and $\rstr{s} : \stringin{r}$, or else $S \err$.
+\subsubsection{Preservation of Security}
+
+Finally, our main result establishes that correctness of $\lambdas$ is preserved under the translation into $\lambdap$. \renewcommand*{\proofname}{Proof}
+
+\begin{thm}[Correctness of Input Sanitation for Translated Terms]\label{thm:main}
+  If $\trden{e} = \iota$ and $\emptyset \vdash e : \stringin{r}$ and $e \sreduces \rstr{s}$ then $\iota \sreduces \tstr{s}$
+  for $s \in \lang{r}$.
 \end{thm}
-\begin{proof}
-By induction on the derivation of $\sctx \vdash S : \psi$. The interesting case
-is S-T-Replace, which requires Lemma C.
-
-\begin{description}
-\item[S-T-Stringin-I:]
-If $S = \rsconcat{S_1}{S_2} : \stringin{r}$ then
-$S \sreduces S$ by S-E-RStr, and $\sctx \vdash S : \psi$ by assumption.
-
-\item[S-T-Concat:] 
-Suppose $S = \rsconcat{S_1}{S_2} : \stringin{r_1 r_2}$.
-By inversion, $\sctx \vdash S_1 : \stringin{r_1}$ and $\sctx \vdash S_1 : \stringin{r_2}$.
-It follows by induction that either $S_1 \err$, $S_2 \err$, 
-or $S_1 \sreduces \rstr{s_1}$ and $S_2 \sreduces \rstr{s_2}$ for some $s_1 \in \lang{r_1}$ and $s_2 \in \lang{r_2}$.
-In the latter case $S \reduces \rstr{s_1s_2}$ by S-E-Concat and $\sctx \vdash \rstr{s_1s_2} : \tstr{r_1r_2}$ by \ref{thm:regexprops}.
-In the former cases, $S \err$.
-
-\item[S-T-Replace:]
-Suppose $S = \rreplace{r}{S_1}{S_2}$ and $\sctx \vdash S : \stringin{r'}$.
-By inversion $\sctx \vdash S_1 : \stringin{r_1}$ and $\sctx \vdash S_2 : \stringin{r_2}$ such that $\lsubst{r}{r_1}{r_2} = r'$.
-By induction, $S_1 \err$, $S_2 \err$ 
-or $S_1 \sreduces \rstr{s_1}$ and $S_2 \sreduces \rstr{s_2}$ such that 
-In the latter case, we know $\lreplace{r}{s_1}{s_2} \in \lang{\lsubst{r}{r_1}{r_2}}$ by Lemma C;
-therefore by S-E-Replace, $S \sreduces \rstr{s}$ such that  $s \in \lang{\lsubst{r}{r_1}{r_2}} = \lang{r'}$.
-So by S-T-String-I, $\rstr{s} : \stringin{r'}$.
-In the former cases, $S \err$.
-
-\item[S-T-Coerce:]
-Suppose $S = \coerce{r}{S_1}$ and $S : \stringin{r}$.
-By inversion, $\sctx \vdash S_1 : \stringin{r'}$.
-By induction, $S_1 \err$ or $S_1 \sreduces \rstr{s}$.
-In the former case $S \err$ by propagation rules.
-In the latter case we have by property 2 of \ref{thm:regexprops} that $s \in \lang{r}$ or else
-$s \not \in \lang{r}$.
-If $s \in \lang{r}$ then $\rstr{s} : \stringin{r}$.
-If $s \not \in \lang{r}$ then $S \err$. 
-\end{description}
-\end{proof}
-
-\begin{thm}
-  Let $P$ be a term in the target language. If $\tctx \vdash P : \theta$ 
-  then $P \sreduces P'$ and $P' : \theta$.
-\end{thm}
-\begin{proof}
-The proof proceeds by induction on the typing relation and is trivial give
-and inversion lemma for the typing relation. \textbf{We can write up this proof if we end up having enough space...}
-\end{proof}
-
-\subsection{Translation Correctness}
-
-We now present the main correctness result.
-
-\begin{thm}
-If $S : \strin{r}$ then there exists a P such that $\trden{s} = P$ and either: 
-\begin{description}
-  \item[(a)] $P \treduces \tstr{s}$ and $S \sreduces \strin{s}$, and $s \in lang{r}$.
-  \item[(b)] $P \err$ and $S \err$.
-  \end{description}
-\end{thm}
-
-\begin{proof}
-The proof proceeds by induction on the typing relation for $S$.
-Throughout the proof, properties from the closure lemma for regular languages
-are necessary; for brevity, we elide these references.
-
-\begin{description}
-\item[S-T-String-I:] 
-Let $S = \rstr{s}$ and suppose $\sctx \vdash \rstr{s} : \stringin{r}$.
-Choose $T = \str{s}$ and note that $\trden{S} = P$ by Tr-String.
-By P-E-String, $P \treduces \str{s}$ and by S-E-String $S \treduces \rstr{s}$.
-Finally, by inversion of S-T-Stringin-I, $s \in \lang{r}$.
-
-\item[S-T-Concat:]
-Let $S = \rconcat{S_1}{S_2}$ and suppose $\sctx \vdash S : \stringin{r_1 r_2}$.
-By inversion, $\sctx \vdash S_1 : \stringin{r_1}$.
-It follows by induction that there exists a $P_1$ such that $\trden{S_1} = P_1$.
-By a similar argument for $S_2$ and $r_2$, there exists a $P_2$ such that $\trden{S_2} = P_2$.
-Choose $P = \concat{P_1}{P_2}$.
-
-We first prove property (a).
-Note that $S_1$ and $P_1$ are well typed (nrf ACTUALLY WE DON'T KNOW THAT $P_1$ IS WELL-TYPED!) and do not result in errors.
-Therefore, $S_1 \sreduces \rstr{s_1}$ and $P_1 \treduces \str{s_1}$ for some $s_1 \in \lang{r_1}$ by theorems 8 and 9 respectively. % TODO-nrf make sure this numbering is correct
-Similarly, $S_2 \sreduces \rstr{s_2}$ and $P_2 \treduces \str{s_2}$ for some $s_2 \in \lang{r_2}$.
-Therefore, $S \sreduces \rstr{s_1 s_2}$ by S-E-Concat and $\concat{P_1}{P_2} \treduces \str{s_1 s_2}$ by P-E-Concat.
-Finally, $s_1s_2 \in \lang{r_1}{r_2}$ by \ref{thm:regexprops}.
-
-Consider property (b).
-If $S_1 \err$ then $P_1 \err$ by induction, and it follows that $S \err$ and $P \err$ by respective error propagation rules.
-Similarly, if $S_2 \err$ then $P_2 \err$ and it follow that $S \err$ and $P \err$ by induction and propagation.
-
-\item[S-T-Replace:]
-Let $S = \rreplace{r}{S_1}{S_2}$ and suppose $\sctx \vdash S : \stringin{r'}$ for some $s$.
-By inversion of S-T-Replace, , $\sctx S_1 : \stringin{r_1}$ and $\sctx : \stringin{r_2}$ 
-such that $\lsubst{r}{r_1}{r_2} = r'$.
-By induction, there exists some $P_1, P_2$ such that $\trden{S_1} = P_1$, $\trden{S_2} = P_2$
-and either (a) or (b) holds.
-
-If (a) holds then $S_1 \sreduces \rstr{s_1}$ and $P_1 \treduces \str{s_1}$ for some $s_1 \in \lang{r_1}$,
-and similarly for $S_2, P_2$ and some $s_2 \in \lang{r_2}$.
-Therefore, by S-E-Replace, $S \sreduces \rstr{s}$ for some $s = \lreplace{r}{s_1}{s_2}$.
-Choose $P = \preplace{r}{s_1}{s_2}$.
-By a similar argument and P-E-Replace, $P \treduces \str{s}$ for some $s = \lreplace{r}{s_1}{s_2}$.
-What remains to be shown is $\lreplace{r}{s_1}{s_2} \in \lang{\lsubst{r}{r_1}{r_2}}$,
-which follows from Leamm D since $s_1 \in r_1$ and $s_2 \in r_2$.
-
-If (b) holds for $S_1$ and $P_1$, then $S \err$ and $P \err$ by propagation rules.
-Similarly, if (b) hols for $S_2$ and $P_2$ then $S \err$ and $P \err$ by propagation rules.
-
-\item[S-T-Coerce:]
-Let $S = \rcoerce{r}{S'}$ and suppose $\sctx \vdash \rcoerce{r}{S} : \stringin{r}$.
-By inverstion $\Psi \vdash S' : \stringin{r'}$ for an arbitrary $r'$.
-By induction there exists a $P'$ such that $\trden{S'} = P'$ and either (a) or (b)
-holds for $S'$ and $P'$.
-
-If (a) holds then $S' \sreduces \rstr{s'}$ and $P' \treduces \str{s'}$ for some $s' \in \lang{r'}$.
-Note that either $s' \in \lang{r}$ or $s' \not \in \lang{r}$ by property 2 of \ref{thm:regexprops}.
-Suppose $s' \in \lang{r}$. Then $\rcoerce{r}{S} \sreduces \rstr{S'}$ by S-E-Coerce.
-Choose $P = \check{\rx{r}}{P'}$ and note that $P \treduces \str{s'}$ by P-E-Coerce.
-Now suppose $s' \not \in \lang{r}$. Then $S \err$ and $P \err$ by P-E-Check-Err
-and S-E-Coerce-Err.
-
-Finally, if (b) holds then $S \err$ and $P \err$ by propagation.
-
-\end{description}
-\end{proof}
-
-% NRF commented out section 2 in order to determine the length of section 1.
-% Replace the pattern ^% with an emptty string in the selection starting at
-% the first empty line of section 2 and ending just before \end{document}.
-
-Papers that needs to be cited in this section:
+\begin{prf}
+  By Theorem \ref{thm:trcorrect}, we have our first conclusion. By Theorem \ref{thm:scorrect} together with the assumption that $e$ is well-typed we have that $s \in \lang{r}$.
+\end{prf}
+
+%\begin{figure}
+%\begin{lstlisting}
+%output of successful compilation
+%\end{lstlisting}
+%\caption{Output of successful compilation.}
+%\end{figure}
+%
+%\begin{figure}
+%\begin{lstlisting}
+%output of failed compilation
+%\end{lstlisting}
+%\caption{Output of failed compilation.}
+%\end{figure}
+%
+
+%\subsection{An Example}\label{whatev}
+
+\begin{figure}[ht]
+\footnotesize
+$\fbox{\inferrule{}{\trden{\sigma}=\tau}}$
+\vspace{-20px}\begin{mathpar}\hspace{20px}
+\inferrule[Tr-T-String]
+{ }
+{ \trden{\stringin{r}} = \str }
+~~~~~~
+\inferrule[Tr-T-Arrow]
+{ \trden{\sigma_1} = \tau_1 \\ \trden{\sigma_2} = \tau_2 }
+{ \trden{\sigma_1 \rightarrow \sigma_2} = \tau_1 \rightarrow \tau_2 }
+\end{mathpar}
+$\fbox{\inferrule{}{\trden{\Psi}=\Theta}}$
+\vspace{-20px}\begin{mathpar}\hspace{20px}
+\inferrule[Tr-T-Context-Emp]
+{ }{\trden{\emptyset} = \emptyset}
+~~~~~~
+\inferrule[Tr-T-Context-Ext]
+{\trden{\Psi}=\Theta\\ \trden{\sigma} = \tau}
+{\trden{\Psi, x : \sigma} = \Theta, x : \tau}
+\end{mathpar}
+  $\fbox{\inferrule{}{ \trden{e} = \iota }}$\vspace{-5px}
+\begin{mathpar}
+\inferrule[Tr-Var]
+{ }
+{ \trden{x} = x}
+
+\inferrule[Tr-Abs]
+{  \trden{e} = \iota }
+{ \trden{ \lambda x.e } = \lambda x . \iota}
+
+\inferrule[Tr-App]
+{ \trden{e_1} = \iota_1 \\  \trden{e_2} = \iota_2}
+{ \trden{e_1(e_2)} = \iota_1(\iota_2)}
+
+\inferrule[Tr-string]
+{ \ }
+{   \trden{\strin{s}} = \tstr{s}}
+
+\inferrule[Tr-Concat]
+{  \trden{e_1} = \iota_1 \\  \trden{e_2} = \iota_2}
+{  \trden{\rsconcat{e_1}{e_2}} = \tconcat{\iota_1}{\iota_2} }
+
+\inferrule[Tr-Case]
+{ \trden{e_1} = \iota_1 \\
+   \trden{e_2} = \iota_2 \\
+   \trden{e_3} = \iota_3
+}
+{ \trden{ \strcase{e_1}{e_2}{x,y.e_3} } = \pstrcase{\iota_1}{\iota_2}{x,y.\iota_3} }
+
+\inferrule[Tr-Replace]
+{   \trden{e_1} = \iota_1 \\   \trden{e_2} = \iota_2 }
+{   \trden{ \sisubst{r}{e_1}{e_2} } = \tsubst{\rx{r}}{\iota_1}{\iota_2} }
+
+\inferrule[Tr-SafeCoerce]
+{  \trden{e} = \iota }
+{   \trden{ \coerce{r'}{e} } = \iota }
+
+\inferrule[Tr-Check]
+{  
+  \trden{e} = \iota \\  
+   \trden{e_1} = \iota_1 \\  
+   \trden{e_2} = \iota_2
+} 
+{ 
+  \trden{ \rcheck{r}{e}{x.e_1}{e_2} } = \tcheck{\rx{r}}{\iota}{(\lambda x . \iota_1)(\iota)}{\iota_2} 
+} 
+\end{mathpar}
+\caption{Translation from source to target.}
+\label{fig:tr}
+\end{figure}
+
+\section{Implementation in \tt{typy}}\label{typy}
+In the previous section, we specified a type system and a translation semantics to a language containing only strings and regular expressions.
+In this section, we take Python to be such a target language.
+Python does not have a static type system, however, so to implement these semantics, we will use \verb|typy|, an extensible type system for Python (being developed by the authors).
+By using \verb|typy|, which leverages Python's quotations and reflection facilities, we can implement these semantics as a library, rather than as a new dialect of the language.
+
+\subsection{Example Usage}
+
+\newcommand{\lstinlinep}[1]{\lstinline[language=Python,basicstyle=\ttfamily\footnotesize]{#1}}
+
+Figure \ref{fig:atexample} demonstrates the use of two type constructors, \verb|fn| and \verb|stringin|, corresponding to the two type constructors of $\lambda_{RS}$.
+We show these as being imported from \verb|typy.std|, the standard library for \verb|typy| (it benefits from no special support from the language itself).
+
+The \verb|fn| type constructor can be used to annotate functions that should be statically checked by \verb|typy|.\footnote{Here, we use argument annotation syntax only available in versions 3.0+ of Python.
+Syntax supporting Python 2.7+ is available, not shown.} The function \verb|sanitize| on lines 3-7, for example,  specifies one argument, \verb|s|, of type \lstinlinep{stringin[r'.*']}.
+Here, the index is a regular expression, written using Python's \emph{raw string notation} as is conventional (in this particular instance, the \verb|r| is not strictly necessary).
+The sanitize function takes an arbitrary string and returns a string without
+double quotes or left and right brackets. Note that the return type need not be specified: \verb|typy| uses a form of \emph{local type inference} \cite{Pierce:2000:LTI:345099.345100}.
+
+In this example, we use an HTML encoding so that the same sanitization function can be used to generate both SQL commands and HTML securely.
+The sanitized string is generated by invoking the \verb|replace| operator, which has the same semantics as $\textsf{rreplace}$ in $\lambda_{RS}$.
+Unlike in the core calculus, it is invoked like a method on \verb|s|.
+The regular expression determining the substrings to be replaced is given as the first argument (as in $\lambda_{RS}$, the only restriction here is that the regular expression must be specified statically.)
+
+\begin{figure}[t]\footnotesize\begin{lstlisting}
+from typy.std import fn, stringin
+
+@fn
+def sanitize(s : stringin[r'.*']):
+  return (s.replace(r'"', '&quot;') 
+           .replace(r'<', '&lt;')
+           .replace(r'>', '&gt;'))
+
+@fn
+def results_query(s : stringin[r'[^"]*']):
+  return 'SELECT * FROM users WHERE name="' + s + '"'
+
+@fn
+def results_div(s : stringin[r'[^<>]*']):
+  return '<div>Results for ' + s + '</div>'
+
+@fn
+def main():
+  input = sanitize(user_input())
+  results = db_execute(results_query(input))
+  return results_div(input) + format(results)
+\end{lstlisting}
+%\vspace{-10px}
+\caption{Regular string types in {\tt typy}}\label{fig:atexample}
+\end{figure}
+
+\begin{figure}
+\begin{lstlisting}
+import re
+
+def sanitize(s):
+   return re.sub(r'"', re.sub(r'<', re.sub(r'>', 
+     s, '&gt;'), '&lt;'), '&quot;')
+
+def results_query(s): 
+  return 'SELECT * FROM users WHERE name="' + s + '"'
+
+def results_div(s):
+  return '<div>Results for ' + s + '</div>'
+
+def main():
+  input = sanitize(user_input())
+  results = db_execute(results_query(input))
+  return results_div(input) + format(results)
+\end{lstlisting}
+%\vspace{-10px}
+\caption{The output of compilation of Figure \ref{fig:atexample} (at the terminal, \texttt{typy figure9.py}, or just-in-time).}
+\label{fig:atexample-out}
+\end{figure}
+
+The functions \verb|results_query| and \verb|results_div| construct a SQL query and an HTML snippet, respectively, by regular string concatenation.
+The argument type annotations serve as a check that sanitation was properly performed.
+In the case of \textsf{results_query}, this specification ensures that user input cannot prematurely be terminated.
+In the case of \textsf{results_div}, this specification ensures that user input does 
+not contain any HTML tags, which is a conservative but effective policy for preventing XSS attacks.
+Note that the type of the surrounding string literals are determined by the type constructor of the function they appear in, \verb|fn| in both cases, which we assume simply chooses \lstinlinep{stringin[r'.*']} (an alternative strategy would be to use the most specific type for the literal, rather than the most general, but this choice is immaterial for this example).
+The addition operator here corresponds to the $\textsf{rconcat}$ operator in $\lambda_{RS}$.
+
+The \verb|main| function invokes the functions just described.
+It begins by passing  user input to \verb|sanitize|, then executing a database query and returning HTML based on this sanitized input.
+The helper functions \verb|user_input| and \verb|db_execute| are not shown but can be assumed standard.
+Importantly, had we mistakenly forgotten to call \verb|sanitize|, the function would not type check (in this case, it is obvious that we did, but lines 14 and 15 would in practice be separated more drastically in the code).
+Moreover, had \verb|sanitize| itself not been implemented correctly (e.g.
+we forgot to strip out quotation marks), then \verb|main| would also not typecheck either.
+
+One somewhat subtle issue here is that the return type of \verb|sanitize| is equivalent to \lstinlinep{stringin[r'[^\"<>]*']}, which is a distinct type from the argument types to \verb|results_query| and \verb|results_div|.
+More specifically, however, it is a ``smaller'' type, in that it could be coerced to these argument types using an operator like $\textsf{rcoerce}$ in $\lambda_{RS}$.
+In our implementation, safe coercions are performed implicitly rather than explicitly.
+Because all regular strings are implemented as strings, this coercion induces no run-time change in representation.
+
+%The result of applying sanitize to input is appended in two functions which
+%construct a safe query (avoiding command injection) and safe HTML output (avoiding Cross-Site Scripting (XSS) attacks).
+%The arguments to the result and output construction functions constitute \emph{specifications}.
+
+%Note that ${\sf input}$ does not actually meet these specifications without additional machinery.
+%The type of ${\sf input}$ is quite large
+%and does not actually equal the specified domains of the query or output construction
+%methods. This mismatch is common -- in fact, nearly universal. Therefore, 
+%our implementation includes a simple subtyping relation between regular expression
+%types.
+
+%This subtyping relation is justified theoretically by the fact that language inclusion is
+%decidable; see \cite{fulton12} for a formal definition of the subtyping relation.
+%Additionally, our extension remains composable because subtyping is defined on a type-by-type basis;
+%see \cite{fulton13} for a discussion of subtyping in Atlang (referred to there as Ace).
+Figure \ref{fig:atexample-out} shows the output of typechecking and translating this code (this can occur either in batch mode at the terminal, generating a new file, or ``just-in-time'' at the first callsite of each function in the dynamically typed portion of the program, not shown).
+
+\subsection{Implementation}
+
+\begin{figure}
+\begin{lstlisting}
+class stringin(typy.Type):
+  def __init__(self, rx):
+    typy.Type.__init__(idx=rx)
+
+  def ana_Str(self, ctx, node):
+    if not in_lang(node.s, self.idx):
+      raise typy.TypeError("...", node)
+
+  def trans_Str(self, ctx, node):
+    return astx.copy(node)
+
+  def syn_BinOp_Add(self, ctx, node):
+    left_t = ctx.syn(node.left)
+    right_t = ctx.syn(node.right)
+    if isinstance(left_t, stringin):
+      left_rx = left_t.idx
+      if isinstance(right_t, stringin):
+        right_rx = right_t.idx
+        return stringin[lconcat(left_rx, right_rx)]
+    raise typy.TypeError("...", node)
+
+  def trans_BinOp_Add(self, ctx, node):
+    return astx.copy(node)
+
+  def syn_Method_replace(self, ctx, node):
+    [rx, exp] = node.args
+    if not isinstance(rx, ast.Str):
+        raise typy.TypeError("...", node)
+    rx = rx.s
+    exp_t = ctx.syn(exp)
+    if not isinstance(exp_t, stringin):
+      raise typy.TypeError("...", node)
+    exp_rx = exp_t.idx
+    return stringin[lreplace(self.idx, rx, exp_rx)]
+
+  def trans_Method_replace(self, ctx, node):
+    return astx.quote(
+        """__import__(re); re.sub(%0, %1, %2)""",
+        astx.Str(s=node.args[0]),
+        astx.copy(node.func.value),
+        astx.copy(node.args[1]))
+
+  # check and strcase omitted
+
+  def check_Coerce(self, ctx, node, other_t):
+    # coercions can only be defined between 
+    # types with the same type constructor, 
+    if rx_sublang(other_t.idx, self.idx):
+      return other_t
+    else: raise typy.TypeError("...", node)
+\end{lstlisting}
+%\vspace{-10px}
+\caption{Implementation of \texttt{stringin} in {\tt typy}.}
+\label{fig:impl}
+
+\label{fig:atimpl}\end{figure}
+The primary unit of extension in \verb|typy| is the \emph{active type constructor}, rather than the abstract syntax as in other work on language extensibility.
+This allows us to implement the entire system as a library in Python and avoid needing to develop new tooling, and also makes it more difficult to create ambiguities between extensions.
+Active type constructors are subclasses of \verb|typy.Type|, and types are instances of these classes.
+The methods of active type constructors control how typechecking and translation proceed for associated operations.
+In Figure \ref{fig:impl}, we show key portions of the implementation of the \verb|stringin| type used in the example above.
+Although a detailed description of the extension mechanism is beyond the scope of this work, we describe the intuitions behind the various methods below.
+
+The constructor, \verb|__init__| in Python, is called when a type is constructed. It simply stores the provided regular expression as the type index by calling the superclass.
+
+When a string literal is being checked against a regular string type, the method \verb|ana_Str| is called.
+It checks that the string is in the language of the regular expression provided as the type index, corresponding to rule \textsc{S-T-Stringin-I} in Section 2.
+The method \verb|trans_Str| is called after typechecking to produce a translation.
+Here, we just copy the original string literal -- regular strings are implemented as strings.
+
+The method \verb|syn_BinOp_Add| synthesizes a type for the string concatenation operation if both arguments are regular strings, consistent with rule \textsc{S-T-Concat}.
+The corresponding method \verb|trans_BinOp_Add| again simply translates the operation directly to string concatenation, consistent with our translation semantics.
+
+The method \verb|syn_Method_replace| synthesizes a type for the ``method-like'' operator \verb|replace|, seen used in our example.
+It ensures that the first argument is a statically known string, using Python's built-in \verb|ast| module, and leverages an implementation of \verb|lreplace|, which computes the appropriate regular expression for the string following replacement, again consistent with our description in Section 2.
+Translation proceeds by using the \verb|re| library built into Python, as can be seen in Figure \ref{fig:atexample-out}.
+
+Code for checked conversions and string decomposition is omitted, but is again consistent with our specification in the previous section.
+Safe coercions are controlled by the \verb|check_Coerce| function, which checks for a sublanguage relationship.
+Here, as in the other methods, failure is indicated by raising an \verb|typy.TypeError| with an appropriate error message and location.
+
+Taken together, we see that the mechanics of extending \verb|typy| with a new type constructor are fairly straightforward: we determine which syntactic forms the operators we specified in our core calculus should correspond to, then directly implement a decision procedure for type synthesis (or, in the case of literal forms, type analysis) in Python.
+The \verb|typy| compiler invokes this logic when working with regular strings.
+The result is a library-based embedding of our semantics into Python.
+
+% We implemented a variation on the type system presented in this paper with two
+% significant differences. first, we only support replacements where $s_2$ is the empty
+% string. Therefore, our implementation respects the system presented in section 2
+% only modulo the definition of \textsf{lreplace}. Seocnd, we use subtyping instead of
+% \textsc{S-*-SafeCoerce}, which decreases inessential verbosity of the language.
+
+% Atlang translates programs using type definitions, which may extend both the
+% static and dynamic semantics of the language. New types are defined as Python
+% classes; figure \ref{fig:impl} contains the source code of our implementation.
+
+% The \textsf{stringin} type has an indexing regular expression \texttt{idx}.
+% Our translation is defined by the \textsf{trans_} methods while the \textsf{syn_}
+% methods define our type checker. Atlang defers type checking and translation
+% to these methods whenever an expression of type \textsf{stringin} is encountered.
+
+\section{Related Work}\label{related}
+
+The input sanitation problem is well-studied.
+There exist a large number of techniques, proposed by both practitioners and researchers, for preventing injection attacks.
+In this section, we explain how our approach to the input sanitation problem differs from each of these approaches.
+Equally important  is our assertion that language extensibility is the right approach for  consideration of language-oriented security mechanisms like the one we described here.
+
+Unlike {frameworks and libraries} provided by languages such as Haskell and Ruby, our type system provides a \emph{static} guarantee that input is always properly sanitized before use.
+Doing so requires reasoning about the operations on regular languages corresponding to standard operations on strings; we are unaware of any production system which contains this form of reasoning.
+Therefore, even where frameworks and libraries provide a viable interface or wrapper around input sanitation (e.g.
+prepared SQL statements), our approach is complementary because it can be used to ensure the correctness of that framework or library itself.
+Furthermore, our approach is more general than database abstraction layers because our mechanism is applicable to all forms of command injection (e.g.
+shell injection or remote file inclusion).
+
+A number of research languages provide static guarantees that a program is free of input sanitation vulnerabilities by never using a string representation at all, but rather desugaring SQL syntax to a safe representation immediately \cite{UrFlowOSDI10}.
+The Wyvern programming language introduced a general framework for writing syntax extensions like this \cite{tsl}.
+Unlike this work, our solution to the input sanitation problem retains a string representation and thus has a very low barrier to adoption.
+Our implementation conservatively extends Python -- a popular language among web developers -- rather than requiring the adoption of a new language entirely.
+We also believe our semantic extensibility approach (as opposed to only syntactic extensibility in Wyvern) is better-positioned for security, where continuously evolving threats might require frequent addition of new semantic  analyses.
+\verb|typy| is particularly well-suited to type system based analyses.
+
+%\begin{itemize}
+%  \item Our system is a light-weight solution to a single class of sanitation vulnerabilities (e.g. we do not address Cross-Site Scripting).
+%  \item Our system is defined as a library in terms of an extensible type system, as opposed to a stand-alone language.
+Instead of introducing new technologies and methodologies for addressing security problems, we provide a light-weight static analysis which complements approaches developers already understand well.
+%  \item Our implementation of the translation is implemented in Python and shares its grammar.
+Since Python is a popular programming language among web developers, the barrier between our research and adopted technologies is lower than for greenfield security-oriented languages.
+%\end{itemize}
+
+% HosoyaPierce2002? 
+Incorporating regular expressions into the type system is not novel. The XDuce system \cite{HosoyaVouillonPierce2000ICFP} checks XML documents against schema using regular expressions. 
+2Similarly, XHaskell \cite{xhaskell} focuses on XML documents.
+We differ from this and related work in at least three ways:
 \begin{itemize}
-  \item Ur/Web OSDI paper
-  \item Jif?
-  \item OWASP
-  \item XDuce and related papers. 
-  \item src?
-  \item Ace or Wyvern paper?
-  \item hotsos?
-  \item Haskell extension paper
-  \item Maybe some popular FOSS libraries/frameworks that do input sanitation?
+  \item Although our static replacement operation is definable in some languages with regular expression types, we are the first to expose this operation and connect the semantics of regular language replacement with the semantics of string replacement via a type theoretic argument.
+  \item The underlying representation of regular strings is guaranteed to be a string, rather than a heavier implementation based on an encoding using functional datatypes.
+  \item We demonstrate that regular expression types are applicable to the web security domain, whereas previous work on regular expression types focused on XML.
+%  \item We work within an extensible type system.
 \end{itemize}
 
+%In conclusion, our contribution is a type system, implemented within an extensible type system, for checking the correctness of input sanitation algorithms.
+
+\section{Future Work}
+
+We believe that this type system extension serves as a useful basis for web-oriented static analysis;
+frameworks and regular expression libraries could be annotated, along with use-sites. We hope to empirically evaluate the feasability of this approach in the future using \verb|typy|. 
+%The use sites of arbitrary strings are usually constrained to a few sections of the application. Assuming
+%regular expressions are used for input sanitation, annotations at these use sites could be sufficient input to a sound and complete static analysis.
+%
+We also believe that extensible programming languages are a promising approach toward incorporating other security
+ analyses into programming languages.
+Construing such analyses as type systems, specifying them rigorously and implementing them within an extensible type system appears to be a promising general technique that the community may wish to emulate.
+%languages might provide developers with a way to incorporate domain-specific security and privacy analyses into their
+%daily development activities. In future work, we hope to bolster this hypothesis by extending our approach in this paper
+%to more realistic settings and by developing new type extensions addressing security and privacy concerns.
+%
+% \section{Conclusion}
+
+% Composable analyses which complement existing approaches constitute a promising approach toward the integration of security concerns into programming languages.
+% In this paper, we presented a system with both of these properties and defined a security-preserving transformation.
+% Unlike other approaches, our solution complements existing, familiar solutions while providing a strong guarantee that traditional library and framework-based approaches are implemented and utilized correctly.
+
+\section*{Acknowledgements}
+We thank the anonymous referees.
+This work was supported by the National Security Agency
+lablet contract \#H98230-14-C-0140 and the National Science Foundation under NSF CNS-1035800.
+%The views and conclusions contained in this document are those of the author and should not be inter- preted as representing the official policies, either expressed or implied, of any sponsoring institution or government.
+Any opinions, findings, and conclusions or recommendations expressed in this publication are those of the author(s) and do not necessarily reflect the views of any sponsoring institution or government.
+\vspace{5px}
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
 
-\newcommand{\F}[1]{\textsf{#1}~}
-\newcommand{\FF}[1]{\textsf{#1}}
-\newcommand{\Q}{\FF{Arg}}
-\newcommand{\xlA}[1]{\lfloor #1 \rfloor_{\lambda}}
-\newcommand{\xA}[1]{$\lfloor #1 \rfloor_{\text{Ace}}$}
-\newcommand{\rtotau}[1]{\lfloor #1 \rfloor}
-
-\newcommand{\tremp}{{\tt tremp}}
-\newcommand{\trdot}{{\tt trdot}}
-\newcommand{\trchar}[1]{{\tt trchar}[#1]}
-\newcommand{\trseq}[2]{{\tt trseq}(#1; #2)}
-\newcommand{\tror}[2]{{\tt tror}(#1; #2)}
-\newcommand{\tlstr}[1]{{\tt tstr}[#1]}
-\begin{table*}[t]
-\centering
-\begin{tabular}{ l l l }
-$r$ & $\xlA{r}$ & \xA{r}\\
-\hline
-$\epsilon$ & $\tremp$ & \verb|""|\\
-$.$ & $\trdot$ & \verb|"."|\\
-$a$ & $\trchar{a}$ & \verb|"|$a$\verb|"|\\
-$r_1 \cdot r_2$ & $\trseq{\xlA{r_1}}{\xlA{r_2}}$ & \xA{r_1}\verb| + |\xA{r_2}\\
-$r_1 + r_2$ & $\tror{\xlA{r_1}}{\xlA{r_2}}$ & \verb|"(" + |\xA{r_1}\verb! + "|" + !\xA{r_2}\verb| + ")"|\\
-\\
-$\psi$ & $\xlA{\psi}$ & \xA{\psi}\\
-\hline
-${\tt string\_in}[r]$ & $\fvar{stringin}[\xlA{r}]$ & \verb|string_in[|\xA{r}\verb|]|\\
-\\
-  $S$ & $\xlA{S}$ & \xA{S} \\
-  \hline
-  ${\tt str\_in}[s]$ & $\FF{intro}[\FF{str}[s]]()$ & \verb|"|$s$\verb|"|\\
-  ${\tt concat}(S_1; S_2)$  & $\xlA{S_1}\cdot\FF{elim}[\tvar{concat}](\xlA{S_2})$ & \xA{S_1}\verb| + |\xA{S_2} \\
-  ${\tt subst}[r](S_1; S_2)$  & $\xlA{S_1}\cdot\FF{elim}[\tvar{subst}~\xlA{r}](\xlA{S_2})$ & \xA{S_1}\verb|.subst(|\xA{r}\verb|, |\xA{S_2}\verb|)|\\
-  ${\tt coerce}[r](S)$  & $\xlA{S}\cdot\FF{elim}[\tvar{coerce}~\xlA{r}]()$ & \xA{S}\verb|.coerce(|\xA{r}\verb|)|
-\end{tabular}
-\caption{Embeddings of the ${\tt string\_in}$ fragment into $\lamAce$ and Ace.}
-\end{table*}
-%
-\newcommand{\atjsynX}[3]{\Gamma \vdash_\fvalCtx #1 \Rightarrow #2 \leadsto #3}
-\newcommand{\atjanaX}[3]{\Gamma \vdash_\fvalCtx #1 \Leftarrow #2 \leadsto #3}
-\newcommand{\atjerrX}[1]{\Gamma \vdash_\fvalCtx #1~ \mathtt{error}}
-\begin{figure*}[t]
-\small
-$\fbox{\inferrule{}{\atjsynX{e}{\tau}{i}}}$~~~~
-$\fbox{\inferrule{}{\atjanaX{e}{\tau}{i}}}$~~~~
-%$\fbox{\inferrule{}{\atjerrX{e}}}$
-\begin{mathpar}
-\inferrule[att-flip]{
-	\atjsynX{e}{\tau}{i}
-}{
-	\atjanaX{e}{\tau}{i}
-}
-
-%\inferrule[att-var]{
-%	x \Rightarrow \tau \in \Gamma
+%\newcommand{\F}[1]{\textsf{#1}~}
+%\newcommand{\FF}[1]{\textsf{#1}}
+%\newcommand{\Q}{\FF{Arg}}
+%\newcommand{\xlA}[1]{\lfloor #1 \rfloor_{\lambda}}
+%\newcommand{\xA}[1]{$\lfloor #1 \rfloor_{\text{Ace}}$}
+%\newcommand{\rtotau}[1]{\lfloor #1 \rfloor}
+%
+%\newcommand{\tremp}{{\tt tremp}}
+%\newcommand{\trdot}{{\tt trdot}}
+%\newcommand{\trchar}[1]{{\tt trchar}[#1]}
+%\newcommand{\trseq}[2]{{\tt trseq}(#1; #2)}
+%\newcommand{\tror}[2]{{\tt tror}(#1; #2)}
+%\newcommand{\tlstr}[1]{{\tt tstr}[#1]}
+%\begin{table*}[t]
+%\centering
+%\begin{tabular}{ l l l }
+%$r$ & $\xlA{r}$ & \xA{r}\\
+%\hline
+%$\epsilon$ & $\tremp$ & \verb|""|\\
+%$.$ & $\trdot$ & \verb|"."|\\
+%$a$ & $\trchar{a}$ & \verb|"|$a$\verb|"|\\
+%$r_1 \cdot r_2$ & $\trseq{\xlA{r_1}}{\xlA{r_2}}$ & \xA{r_1}\verb| + |\xA{r_2}\\
+%$r_1 + r_2$ & $\tror{\xlA{r_1}}{\xlA{r_2}}$ & \verb|"(" + |\xA{r_1}\verb! + "|" + !\xA{r_2}\verb| + ")"|\\
+%\\
+%$\sigma$ & $\xlA{\sigma}$ & \xA{\sigma}\\
+%\hline
+%${\tt string\_in}[r]$ & $\fvar{stringin}[\xlA{r}]$ & \verb|stringin[|\xA{r}\verb|]|\\
+%\\
+%  $S$ & $\xlA{S}$ & \xA{S} \\
+%  \hline
+%  ${\tt str\_in}[s]$ & $\FF{intro}[\FF{str}[s]]()$ & \verb|"|$s$\verb|"|\\
+%  ${\tt concat}(S_1; S_2)$  & $\xlA{S_1}\cdot\FF{elim}[\tvar{concat}](\xlA{S_2})$ & \xA{S_1}\verb| + |\xA{S_2} \\
+%  ${\tt subst}[r](S_1; S_2)$  & $\xlA{S_1}\cdot\FF{elim}[\tvar{subst}~\xlA{r}](\xlA{S_2})$ & \xA{S_1}\verb|.subst(|\xA{r}\verb|, |\xA{S_2}\verb|)|\\
+%  ${\tt coerce}[r](S)$  & $\xlA{S}\cdot\FF{elim}[\tvar{coerce}~\xlA{r}]()$ & \xA{S}\verb|.coerce(|\xA{r}\verb|)|
+%\end{tabular}
+%\caption{Embeddings of the ${\tt string\_in}$ fragment into $\lamAce$ and Ace.}
+%\end{table*}
+%%
+%\newcommand{\atjsynX}[3]{\Gamma \vdash_\fvalCtx #1 \Rightarrow #2 \leadsto #3}
+%\newcommand{\atjanaX}[3]{\Gamma \vdash_\fvalCtx #1 \Leftarrow #2 \leadsto #3}
+%\newcommand{\atjerrX}[1]{\Gamma \vdash_\fvalCtx #1~ \mathtt{error}}
+%\begin{figure*}[t]
+%\small
+%$\fbox{\inferrule{}{\atjsynX{e}{\tau}{i}}}$~~~~
+%$\fbox{\inferrule{}{\atjanaX{e}{\tau}{i}}}$~~~~
+%%$\fbox{\inferrule{}{\atjerrX{e}}}$
+%\begin{mathpar}
+%\inferrule[att-flip]{
+%	\atjsynX{e}{\tau}{i}
 %}{
-%	\atjsynX{x}{\tau}{x}
+%	\atjanaX{e}{\tau}{i}
 %}
 %
-%\inferrule[att-asc]{
-%    \tau \Downarrow_\fvalCtx \tau'\\
-%	\atjanaX{e}{\tau'}{i}
-%}{
-%	\atjsynX{e : \tau}{\tau'}{i}
-%}
-%
-%\inferrule[att-let-syn]{
-%	\atjsynX{e_1}{\tau_1}{i_1}\\
-%	\Gamma, x \Rightarrow \tau_1 \vdash_\fvalCtx e_2 \Rightarrow \tau_2 \leadsto i_2\\
-%	\trepof{\tau_1} \Downarrow_\fvalCtx \titype{\sigma_1}
-%}{
-%	\atjsynX{\F{let}x = e_1~\F{in}e_2}{\tau_2}{(\ilam{x}{\sigma_1}{i_2})~i_1}
-%}
-%
-%\inferrule[att-lam-ana]{
-%	\Gamma, x \Rightarrow \tau_1 \vdash_\fvalCtx e \Leftarrow \tau_2 \leadsto i\\
-%	\trepof{\tau_1} \Downarrow_\fvalCtx \titype{\sigma_1}
-%}{
-%	\atjanaX{\lambda x.e}{\ttype{arrow}{(\tau_1, \tau_2)}}{\ilam{x}{\sigma_1}{i}}
-%}
-%
-%\inferrule[att-lam]{
-%	\tau_1 \Downarrow_\fvalCtx \tau_1'\\
-%		\trepof{\tau_1'} \Downarrow_\fvalCtx \titype{\sigma}\\\\
-%	\Gamma, x \Rightarrow \tau_1' \vdash_\fvalCtx e \Rightarrow \tau_2 \leadsto i
-%%	i \hookrightarrow_\fvalCtx i'\\
-%%		\sigma \hookrightarrow_\fvalCtx \sigma'
-%%	\ddbar{\fvar{Arrow}}{\fvalCtx}{\trepof{\tau_1'}}{\sbar_1}\\
-%	%\delfromtau{$\Xi_0$}{\fvalCtx}{\tau_1'}{\sabs}\\\\
-%}{
-%	\atjsynX{\elam{x}{\tau_1}{e}}{\ttype{arrow}{(\tau_1', \tau_2)}}{\ilam{x}{\sigma'}{i}}
-%}
-%
-\inferrule[att-intro-ana]{
-	\vdash_\fvalCtx \FF{iana}(\fvar{tycon})=\taudef\\
-	\taudef~\taut{opidx}~\taut{tyidx}~((\Gamma; e_1)? :: \ldots :: (\Gamma; e_n)? :: []) \Downarrow_\fvalCtx \titerm{i}\\\\
-%	\trepof{\ttype{tycon}{\tauidx'}} \Downarrow_\fvalCtx \titype{\sigma}\\
-		\trepof{\ttype{tycon}{\taut{tyidx}}} \Downarrow_\fvalCtx \titype{\sigma}\\
-		\Gamma \vdash_\fvalCtx i : \sigma
-}{
-	\atjanaX{\FF{intro}[\taut{opidx}](e_1; \ldots; e_n)}{\ttype{tycon}{\taut{tyidx}}}{i}
-}
-
-%\inferrule[att-i-asc-ty]{
-%	\atjanaX{I}{\tau}{i}
-%}{
-%	\atjsynX{I : \tau}{\tau}{i}
-%}
-%
-%\inferrule[att-i-asc-tycon]{
-%	\vdash_\fvalCtx \FF{isyn}(\fvar{tycon})=\taudef\\\\
-%	\taudef~\tauidx~((\Gamma; e_1)? {::}{\ldots}{::}(\Gamma; e_n)? {::} []) \Downarrow_\fvalCtx \tden{\titerm{i}}{\ttype{tycon}{\tauidx'}}\\
-%			\trepof{\tau_1'} \Downarrow_\fvalCtx \titype{\sigma}\\
+%%\inferrule[att-var]{
+%%	x \Rightarrow \tau \in \Gamma
+%%}{
+%%	\atjsynX{x}{\tau}{x}
+%%}
+%%
+%%\inferrule[att-asc]{
+%%    \tau \Downarrow_\fvalCtx \tau'\\
+%%	\atjanaX{e}{\tau'}{i}
+%%}{
+%%	\atjsynX{e : \tau}{\tau'}{i}
+%%}
+%%
+%%\inferrule[att-let-syn]{
+%%	\atjsynX{e_1}{\tau_1}{i_1}\\
+%%	\Gamma, x \Rightarrow \tau_1 \vdash_\fvalCtx e_2 \Rightarrow \tau_2 \leadsto i_2\\
+%%	\trepof{\tau_1} \Downarrow_\fvalCtx \titype{\sigma_1}
+%%}{
+%%	\atjsynX{\F{let}x = e_1~\F{in}e_2}{\tau_2}{(\ilam{x}{\sigma_1}{i_2})~i_1}
+%%}
+%%
+%%\inferrule[att-lam-ana]{
+%%	\Gamma, x \Rightarrow \tau_1 \vdash_\fvalCtx e \Leftarrow \tau_2 \leadsto i\\
+%%	\trepof{\tau_1} \Downarrow_\fvalCtx \titype{\sigma_1}
+%%}{
+%%	\atjanaX{\lambda x.e}{\ttype{arrow}{(\tau_1, \tau_2)}}{\ilam{x}{\sigma_1}{i}}
+%%}
+%%
+%%\inferrule[att-lam]{
+%%	\tau_1 \Downarrow_\fvalCtx \tau_1'\\
+%%		\trepof{\tau_1'} \Downarrow_\fvalCtx \titype{\sigma}\\\\
+%%	\Gamma, x \Rightarrow \tau_1' \vdash_\fvalCtx e \Rightarrow \tau_2 \leadsto i
+%%%	i \hookrightarrow_\fvalCtx i'\\
+%%%		\sigma \hookrightarrow_\fvalCtx \sigma'
+%%%	\ddbar{\fvar{Arrow}}{\fvalCtx}{\trepof{\tau_1'}}{\sbar_1}\\
+%%	%\delfromtau{$\Xi_0$}{\fvalCtx}{\tau_1'}{\sabs}\\\\
+%%}{
+%%	\atjsynX{\elam{x}{\tau_1}{e}}{\ttype{arrow}{(\tau_1', \tau_2)}}{\ilam{x}{\sigma'}{i}}
+%%}
+%%
+%\inferrule[att-intro-ana]{
+%	\vdash_\fvalCtx \FF{iana}(\fvar{tycon})=\taudef\\
+%	\taudef~\taut{opidx}~\taut{tyidx}~((\Gamma; e_1)? :: \ldots :: (\Gamma; e_n)? :: []) \Downarrow_\fvalCtx \titerm{i}\\\\
+%%	\trepof{\ttype{tycon}{\tauidx'}} \Downarrow_\fvalCtx \titype{\sigma}\\
+%		\trepof{\ttype{tycon}{\taut{tyidx}}} \Downarrow_\fvalCtx \titype{\sigma}\\
 %		\Gamma \vdash_\fvalCtx i : \sigma
 %}{
-%	\atjsynX{\FF{intro}[\tauidx](e_1; \ldots; e_n)] :: \fvar{tycon}}{\ttype{tycon}{\tauidx'}}{i'}
+%	\atjanaX{\FF{intro}[\taut{opidx}](e_1; \ldots; e_n)}{\ttype{tycon}{\taut{tyidx}}}{i}
 %}
 %
-\inferrule[att-elim-syn]{
-	\atjsynX{e}{\ttype{tycon}{\taut{tyidx}}}{i}\\
-	\vdash_\fvalCtx \FF{esyn}(\fvar{tycon})=\taudef\\\\
-	\taudef~\taut{opidx}~\taut{tyidx}~\titerm{i}~((\Gamma; e)? :: (\Gamma; e_1)? :: \ldots :: (\Gamma; e_n)? :: []) \Downarrow_\fvalCtx (\tau, \titerm{i'})\\\\
-			\trepof{\tau} \Downarrow_\fvalCtx \titype{\sigma}\\
-	\Gamma \vdash_\fvalCtx i' : \sigma
-}{
-	\atjsynX{e\cdot\FF{elim}[\taut{opidx}](e_1; \ldots; e_n)}{\tau}{i'}
-}
-\end{mathpar}
-%\vspace{-10px}
-\caption{The bidirectional active typechecking and translation judgements.}
-\label{atj}
-\end{figure*}
-\newcommand{\tlevalX}[2]{#1 \Downarrow_\fvalCtx #2}
-\begin{figure*}[t]
-\small
-$\fbox{\inferrule{}{\tau \Downarrow_\fvalCtx \tau'}}$~~~~
-\begin{mathpar}
-\inferrule[trstr-eval]{ }{\tlevalX{\tlstr{s}}{\tlstr{s}}}
-
-\inferrule[tremp-eval]{ }{\tremp \Downarrow_\fvalCtx \tremp}
-
-\cdots
-
-\inferrule[tror-eval]{
-	\tlevalX{\tau_1}{\tau_1'}\\
-	\tlevalX{\tau_2}{\tau_2'}
-}{
-	\tlevalX{\tror{\tau_1}{\tau_2}}{\tror{\tau_1'}{\tau_2'}}
-}
-
-trmatch
-
-trlsubst
-
-trsublang
-
-%\inferrule[repof]{
-%	\tau \Downarrow_\fvalCtx \ttype{tycon}{\tauidx}\\
-%	\vdash_\fvalCtx \FF{rep}(\fvar{tycon}) = \taurep\\
-%	\taurep~\tauidx \Downarrow_\fvalCtx \titype{\sigma}
+%%\inferrule[att-i-asc-ty]{
+%%	\atjanaX{I}{\tau}{i}
+%%}{
+%%	\atjsynX{I : \tau}{\tau}{i}
+%%}
+%%
+%%\inferrule[att-i-asc-tycon]{
+%%	\vdash_\fvalCtx \FF{isyn}(\fvar{tycon})=\taudef\\\\
+%%	\taudef~\tauidx~((\Gamma; e_1)? {::}{\ldots}{::}(\Gamma; e_n)? {::} []) \Downarrow_\fvalCtx \tden{\titerm{i}}{\ttype{tycon}{\tauidx'}}\\
+%%			\trepof{\tau_1'} \Downarrow_\fvalCtx \titype{\sigma}\\
+%%	\Gamma \vdash_\fvalCtx i : \sigma
+%%}{
+%%	\atjsynX{\FF{intro}[\tauidx](e_1; \ldots; e_n)] :: \fvar{tycon}}{\ttype{tycon}{\tauidx'}}{i'}
+%%}
+%%
+%\inferrule[att-elim-syn]{
+%	\atjsynX{e}{\ttype{tycon}{\taut{tyidx}}}{i}\\
+%	\vdash_\fvalCtx \FF{esyn}(\fvar{tycon})=\taudef\\\\
+%	\taudef~\taut{opidx}~\taut{tyidx}~\titerm{i}~((\Gamma; e)? :: (\Gamma; e_1)? :: \ldots :: (\Gamma; e_n)? :: []) \Downarrow_\fvalCtx (\tau, \titerm{i'})\\\\
+%			\trepof{\tau} \Downarrow_\fvalCtx \titype{\sigma}\\
+%	\Gamma \vdash_\fvalCtx i' : \sigma
 %}{
-%	\FF{repof}(\tau) \Downarrow_\fvalCtx \titype{\sigma}
+%	\atjsynX{e\cdot\FF{elim}[\taut{opidx}](e_1; \ldots; e_n)}{\tau}{i'}
 %}
-%
-\inferrule[syn]{
-    \tau \Downarrow_\fvalCtx (\Gamma; e)?\\
-	\atjsynX{e}{\tau}{\iota}\\\\
-    [\tau/\tvar{t}_{ty}, \titerm{\iota}/\tvar{t}_{trans}]\tau_2 \Downarrow_\fvalCtx \tau_2'
-}{
-	\FF{syn}(\tau_1; \tvar{t}_{ty}, \tvar{t}_{trans}.\tau_2) \Downarrow_\fvalCtx \tau_2'
-}
-
-\inferrule[ana]{
-	\tau_1 \Downarrow_\fvalCtx (\Gamma; e)?\\
-	\tau_2 \Downarrow_\fvalCtx \tau_2'\\
-	\atjanaX{e}{\tau_2'}{\iota}\\\\
-	[\titerm{\iota}/\tvar{t}_{trans}]\tau_3 \Downarrow_\fvalCtx \tau_3'
-}{
-	\FF{ana}(\tau_1; \tau_2; \tvar{t}_{trans}.\tau_3) \Downarrow_\fvalCtx \tau_3'
-}
-
-TODO: add rx, str stuff
-\end{mathpar}
-\caption{\small Normalization semantics for the type-level language.
+%\end{mathpar}
+%%\vspace{-10px}
+%\caption{The bidirectional active typechecking and translation judgements.}
+%\label{atj}
+%\end{figure*}
+%\newcommand{\tlevalX}[2]{#1 \Downarrow_\fvalCtx #2}
+%\begin{figure*}[t]
+%\small
+%$\fbox{\inferrule{}{\tau \Downarrow_\fvalCtx \tau'}}$~~~~
+%\begin{mathpar}
+%\inferrule[trstr-eval]{ }{\tlevalX{\tlstr{s}}{\tlstr{s}}}
+%
+%\inferrule[tremp-eval]{ }{\tremp \Downarrow_\fvalCtx \tremp}
+%
+%\cdots
+%
+%\inferrule[tror-eval]{
+%	\tlevalX{\tau_1}{\tau_1'}\\
+%	\tlevalX{\tau_2}{\tau_2'}
+%}{
+%	\tlevalX{\tror{\tau_1}{\tau_2}}{\tror{\tau_1'}{\tau_2'}}
+%}
+%
+%trmatch
+%
+%trlsubst
+%
+%trsublang
+%
+%
+%%\inferrule[repof]{
+%%	\tau \Downarrow_\fvalCtx \ttype{tycon}{\tauidx}\\
+%%	\vdash_\fvalCtx \FF{rep}(\fvar{tycon}) = \taurep\\
+%%	\taurep~\tauidx \Downarrow_\fvalCtx \titype{\sigma}
+%%}{
+%%	\FF{repof}(\tau) \Downarrow_\fvalCtx \titype{\sigma}
+%%}
+%%
+%\inferrule[syn]{
+%    \tau \Downarrow_\fvalCtx (\Gamma; e)?\\
+%	\atjsynX{e}{\tau}{\iota}\\\\
+%    [\tau/\tvar{t}_{ty}, \titerm{\iota}/\tvar{t}_{trans}]\tau_2 \Downarrow_\fvalCtx \tau_2'
+%}{
+%	\FF{syn}(\tau_1; \tvar{t}_{ty}, \tvar{t}_{trans}.\tau_2) \Downarrow_\fvalCtx \tau_2'
+%}
+%
+%\inferrule[ana]{
+%	\tau_1 \Downarrow_\fvalCtx (\Gamma; e)?\\
+%	\tau_2 \Downarrow_\fvalCtx \tau_2'\\
+%	\atjanaX{e}{\tau_2'}{\iota}\\\\
+%	[\titerm{\iota}/\tvar{t}_{trans}]\tau_3 \Downarrow_\fvalCtx \tau_3'
+%}{
+%	\FF{ana}(\tau_1; \tau_2; \tvar{t}_{trans}.\tau_3) \Downarrow_\fvalCtx \tau_3'
+%}
+%
+%TODO: add rx, str stuff
+%\end{mathpar}
+%\caption{\small Normalization semantics for the type-level language.
 Missing rules (including error propagation rules and normalization of quoted internal terms and types) are unsurprising and will be given later.}
-\label{tleval}
-\end{figure*}
-\begin{figure}[t]
-\small\begin{flalign}
-& \F{tycon}\fvar{stringin}~\F{of}\FF{R}~\{\\
-& \quad \F{iana}\{\tlam{opidx}{\FF{String}}{
-	\tlam{tyidx}{\FF{R}}{
-	\tlam{a}{\klist{\Q}}{\\
-& \quad\quad \tvar{arity0}~\tvar{a}~(\tvar{check}~\tvar{opidx}~\tvar{tyidx}~\titerm{\iup{\tvar{opidx}}})
-	}}
-}\}\\
-& \quad \F{esyn}\{\tlam{opidx}{\kunit+(\FF{R}+\FF{R})}{
-	\tlam{a}{\klist{\Q}}{\\
-& \quad\quad \tsumcase{\tvar{opidx}}{\_}{\\
-& \quad\quad\quad \tvar{arity2}~\tvar{a}~\tlam{a1}{\Q}{\tlam{a2}{\Q}{\\
-& \quad\quad\quad\quad \tvar{rsyn}~\tvar{a1}~\tlam{r1}{\FF{R}}{\tlam{i1}{\kITerm}{~}}\\
-& \quad\quad\quad\quad \tvar{rsyn}~\tvar{a2}~\tlam{r2}{\FF{R}}{\tlam{i2}{\kITerm}{~}}\\
-& \quad\quad\quad\quad\quad (\ttype{stringin}{\FF{rseq}(\tvar{r1}; \tvar{r2})}, \\
-& \quad\quad\quad\quad\quad\titerm{{\tt iconcat}(\iup{\tvar{i1}}; \iup{\tvar{i2}})})
-}}\\
-& \quad\quad}{opidx'}{d}}
-}\}
-\};\\
-& \F{def}\tvar{concat} = \tinl{\kunit, \FF{R}+\FF{R}}{\tunit};\\
-& \F{def}\tvar{subst} = \tlam{r}{\FF{R}}{\tinr{\kunit, \FF{R}+\FF{R}}{\tinl{\FF{R},\FF{R}}{\tvar{r}}}};\\
-& \F{def}\tvar{coerce} = \tlam{r}{\FF{R}}{\tinr{\kunit, \FF{R}+\FF{R}}{\tinr{\FF{R},\FF{R}}{\tvar{r}}}}
-\end{flalign}
-\caption{Definition of $\phi_S$, which enables the embedding of fragment $S$ into $\lamAce$.}
-\end{figure}
-%\begin{figure}
-%\[
-%\begin{array}{lcl}
-%% & & Definition & Kind\\
-%\tvar{concat} & := & \tinl{\kunit, \FF{R}+\FF{R}}{\tunit}\\%& \kunit + (R + R)\\
-%\tvar{replace} & := & \tlam{r}{\FF{R}}{\tinr{\kunit, \FF{R}+\FF{R}}{\tinl{\FF{R},\FF{R}}{\tvar{r}}}}\\% & \karrow{a}{b}\\
-%\tvar{coerce} & := & \tlam{r}{\FF{R}}{\tinr{\kunit, \FF{R}+\FF{R}}{\tinr{\FF{R},\FF{R}}{\tvar{r}}}}\\
-%\end{array}
-%\]
-%\caption{Definitions}
+%\label{tleval}
+%\end{figure*}
+%\begin{figure}[t]
+%\small\begin{flalign}
+%& \F{tycon}\fvar{stringin}~\F{of}\FF{R}~\{\\
+%& \quad \F{iana}\{\tlam{opidx}{\FF{String}}{
+%	\tlam{tyidx}{\FF{R}}{
+%	\tlam{a}{\klist{\Q}}{\\
+%& \quad\quad \tvar{arity0}~\tvar{a}~(\tvar{check}~\tvar{opidx}~\tvar{tyidx}~\titerm{\iup{\tvar{opidx}}})
+%	}}
+%}\}\\
+%& \quad \F{esyn}\{\tlam{opidx}{\kunit+(\FF{R}+\FF{R})}{
+%	\tlam{a}{\klist{\Q}}{\\
+%& \quad\quad \tsumcase{\tvar{opidx}}{\_}{\\
+%& \quad\quad\quad \tvar{arity2}~\tvar{a}~\tlam{a1}{\Q}{\tlam{a2}{\Q}{\\
+%& \quad\quad\quad\quad \tvar{rsyn}~\tvar{a1}~\tlam{r1}{\FF{R}}{\tlam{i1}{\kITerm}{~}}\\
+%& \quad\quad\quad\quad \tvar{rsyn}~\tvar{a2}~\tlam{r2}{\FF{R}}{\tlam{i2}{\kITerm}{~}}\\
+%& \quad\quad\quad\quad\quad (\ttype{stringin}{\FF{rseq}(\tvar{r1}; \tvar{r2})}, \\
+%& \quad\quad\quad\quad\quad\titerm{{\tt iconcat}(\iup{\tvar{i1}}; \iup{\tvar{i2}})})
+%}}\\
+%& \quad\quad}{opidx'}{d}}
+%}\}
+%\};\\
+%& \F{def}\tvar{concat} = \tinl{\kunit, \FF{R}+\FF{R}}{\tunit};\\
+%& \F{def}\tvar{subst} = \tlam{r}{\FF{R}}{\tinr{\kunit, \FF{R}+\FF{R}}{\tinl{\FF{R},\FF{R}}{\tvar{r}}}};\\
+%& \F{def}\tvar{coerce} = \tlam{r}{\FF{R}}{\tinr{\kunit, \FF{R}+\FF{R}}{\tinr{\FF{R},\FF{R}}{\tvar{r}}}}
+%\end{flalign}
+%\caption{Definition of $\phi_S$, which enables the embedding of fragment $S$ into $\lamAce$.}
 %\end{figure}
-
-\subsection{Background: Ace}
-TODO: Make a TR out of the OOPSLA submission.
-\subsection{Explicit Conversions}
-\subsection{Adding Subtyping to Ace}
-\subsection{Theory}
-
-\section{Related Work}
-
-\section{Discussion}
-
+%%\begin{figure}
+%%\[
+%%\begin{array}{lcl}
+%%% & & Definition & Kind\\
+%%\tvar{concat} & := & \tinl{\kunit, \FF{R}+\FF{R}}{\tunit}\\%& \kunit + (R + R)\\
+%%\tvar{replace} & := & \tlam{r}{\FF{R}}{\tinr{\kunit, \FF{R}+\FF{R}}{\tinl{\FF{R},\FF{R}}{\tvar{r}}}}\\% & \karrow{a}{b}\\
+%%\tvar{coerce} & := & \tlam{r}{\FF{R}}{\tinr{\kunit, \FF{R}+\FF{R}}{\tinr{\FF{R},\FF{R}}{\tvar{r}}}}\\
+%%\end{array}
+%%\]
+%%\caption{Definitions}
+%%\end{figure}
+%
+%\subsection{Background: Ace}
+%TODO: Make a TR out of the OOPSLA submission.
+%\subsection{Explicit Conversions}
+%\subsection{Adding Subtyping to Ace}
+%\subsection{Theory}
+%
+%\section{Related Work}
+%
+%\section{Discussion}
+%
+%
 \bibliographystyle{abbrv}
-
+%\renewcommand*{\bibfont}{\small}
+\setlength{\bibsep}{0pt plus .20pt minus .25pt}
+{
+\vspace*{-1.5mm}
+\bibliography{research}
+}
 % The bibliography should be embedded for final submission.
 
-\bibliography{../research}
-%\softraggedright
-%P. Q. Smith, and X. Y. Jones. ...reference text...
-
+%\begin{thebibliography}
+%
+%\bibitem{lamport94}
+%Leslie Lamport,
+%\emph{\LaTeX: a document preparation system}.
+%Addison Wesley, Massachusetts,
+%2nd edition,
+%1994.
+%
+%\end{thebibliography}
 \end{document}

### 2014-12-14 sanitation-psp14/tr/strings_chapter.tex
diagrams.
nop.
Added example of new case for => direction of semantic alignment

@@
 
+% bookmark1
+\begin{trlem} \label{lem:inductive}
+If $e \sreduces v$ and $e \mapsto e'$ then $e' \sreduces v$. 
+\end{trlem}
+\begin{proof}
+By induction on the structure of e.
+\begin{itemize}[label=$ $,itemsep=1ex]
+
+\item \textbf{Case $e = e_1(e_2)$}.
+The only applicable rule is S-E-App, so $e_1 \sreduces \lambda x . e_3$
+and $e_2 \sreduces v_2$ such that $[v_2/x] e_3 \sreduces v$.
+
+The term $e = e_1(e_2)$ may step by three rules.
+
+First, if $e$ steps by L-E-AppLeft then $e_1(e_2) \mapsto e_1'(e_2)$.
+By induction, $e_1' \sreduces \lambda x . e_3$.
+By S-E-App, $e_1'(e_2) \sreduces \lambda x . e_3$.
+
+Second, if $e$ steps by L-E-AppRight then $e_1(e_2) \mapsto e_1(e_2')$.
+By induction, $e_2' \sreduces v_2$.
+By S-E-App, $e_1(e_2') \sreduces \lambda x . e_3$.
+
+Third, if $e$ steps by L-E-AppAbs then $e = (\lambda x . e_3)(v_2) \mapsto [v_2 / x]e_3$.
+By induction, $e = (\lambda x . e')v_2 \sreduces v$.
+\end{itemize}
+\end{proof}
+
 \begin{trthm}[Semantic Correspondence for $\lambdas$ (Part I)]
@@
 
+% bookmark2
 % Application case.
 \item \textbf{Case $e=e_1(e_2)$}.
-The only applicable rule is S-E-App. By inversion, we establish that the following:
+The only applicable rule is S-E-App. By inversion:
 \begin{align*}
@@
 \end{align*}
+
+If $e_1 = \lambda x . e_1'$ and $e_2 = v_2$ (henceforth the reflexive case)
+then $e \mapsto [v_2 / x] e_1'$ and the conclusion follows by RT-Trans.
+
+If $e_1 \mapsto \lambda x . e_1'$ then $e_1(e_2) \mapsto (\lambda x . e_1')(e_2) \mapsto [e_2/x]e_1'$ and the conclusion follows by two applications
+of RT-Trans.
+
+If $e_1 \mapsto^k \lambda x . e_2'$ then $e_1 \mapsto e'$ so $e_1(e_2) \mapsto e'(e_2)$ and by Lemma~\ref{lem:inductive} $e'(e_2) \mapsto^* v$.
+So $e_1(e_2) \mapsto e'(e_2) \mapsto^* v$; it follows by RT-Trans that $e_1(e_2) \mapsto v$.
+
+\todo{not modified below this line...}
 
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-\section{String Substitution and Language Replacement}
-
-\subsection{The Trivial Definition}
-
-\subsection{An Automaton Construction}
-
-Insert Automaton stuff...
-
-\subsection{Toward a Precise Definition}
-
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+\newpage
 
@@
 <$e$> ::= 
-      $x$ | $\lambda x . e$ | $e(e) $\hfill source terms \alt 
-      $\strin{s}$ | $\rsconcat{e}{e}$ | $\strcase{e}{e}{x,y.e}$ \hfill $s \in \Sigma^{*}$ \alt
+      $x$ | $v$ \hfill source terms \alt 
+      $\rsconcat{e}{e}$ | $\strcase{e}{e}{x,y.e}$ \hfill $s \in \Sigma^{*}$ \alt
       $\sisubst{r}{e}{e}$ | $\rcoerce{r}{e}$ | $\rcheck{r}{e}{x.e}{e}$
@@
 
-<$\iota$> ::= $x$ | $\lambda x . \iota$ | $\iota(\iota)$ \hfill target terms \alt
-$\tstr{s}$ | $\tconcat{\iota}{\iota}$ | $\pstrcase{\iota}{\iota}{x,y.\iota}$ \alt
+<$\iota$> ::= $x$ | $\dot{v}$ \hfill target terms \alt
+$\tconcat{\iota}{\iota}$ | $\pstrcase{\iota}{\iota}{x,y.\iota}$ \alt
   $\rx{r}$ | $\preplace{\iota}{\iota}{\iota}$ | $\tcheck{\iota}{\iota}{\iota}{\iota}$
 
-  <$\dot{v}$> ::= $\lambda x . \iota$ | \tstr{s} | \rx{r} \hfill target values
+  <$\dot{v}$> ::= $\lambda x . \iota$ | $\tstr{s}$ | $\rx{r}$ \hfill target values
 
@@
 \inferrule[S-T-Replace]
-{\sctx \vdash e_1 : \stringin{r_1} \\ \sctx \vdash e_2 : \stringin{r_2} \\\\ \lreplace{r}{r_1}{r_2} = r'}
-{\sctx \vdash \sisubst{r}{e_1}{e_2} : \stringin{r'}}
+{\sctx \vdash e_1 : \stringin{r_1} \\ \sctx \vdash e_2 : \stringin{r_2} }
+{\sctx \vdash \sisubst{r}{e_1}{e_2} : \stringin{\lreplace{r}{r_1}{r_2}}}
 
@@
   \small
-$\fbox{\inferrule{}{e \ \val}}$
-
-\begin{mathpar}
-\inferrule[L-Val]
-{  }
-{\lambda x:\tau.t \ \val}
-\end{mathpar}
-
 $\fbox{\inferrule{}{e \mapsto e}}$
@@
 {e_1 \mapsto e_1'}
-{e_1 e_2 \mapsto e_1' e_2}
+{e_1(e_2) \mapsto e_1'(e_2)}
 
@@
 {e_2 \mapsto e_2'}
-{e_1 e_2 \mapsto e_1 e_2'}
+{v_1 \mapsto v_1}
 
@@
 \inferrule[RT-Trans]
-{ e \mapsto^* e' \\ e' \mapsto^* e'' }
+{ e \mapsto^* e' \\ e' \mapsto e'' }
 { e \mapsto^* e'' }
-
-\inferrule[RT-Step\footnote{This is a special case of RT-Trans which we distinguish in order to write clearer proofs.}]
-{ e \mapsto^* e' \\ e' \mapsto v }
-{ e \mapsto^* v }
 
@@
   \small
+$\fbox{\inferrule{}{e \mapsto e}}$ (Contiues figure 6)
 %$\fbox{\inferrule{}{e \mapsto e}} \  \fbox{ \inferrule{}{v \ \val}}$
 \begin{mathpar}
-\inferrule[SS-E-RStr]
-{ \ }
-{\strin{s} \ \val}
-
 \inferrule[SS-E-Concat-Left]
@@
 {e_2 \mapsto e_2'}
-{\rsconcat{e_1}{e_2} \mapsto \rsconcat{e_1}{e_2'}}
+{\rsconcat{v_1}{e_2} \mapsto \rsconcat{v_1}{e_2'}}
 
@@
 
-\inferrule[SS-E-Case-Right]
-{
-  e_2 \mapsto e_2'
-}
-{
-  \strcase{e_1}{e_2}{x,y.e_3} \mapsto \strcase{e_1}{e_2'}{x,y.e_3}
-}
-
 \inferrule[SS-E-Case-$\epsilon$-val]
@@
 {e_1 \mapsto e_1'}
-{ \sisubst{r}{e_1}{e_2} \mapsto \sisubst{r}{e_1'}{e_2}}
+{ \sisubst{r}{v_1}{e_2} \mapsto \sisubst{r}{v_1'}{e_2}}
 
@@
 
-\inferrule[SS-E-Check-StepRight]
-{e_2 \mapsto e_2'}
-{\rcheck{r}{e}{x.e_1}{e_2} \mapsto \rcheck{r}{e}{x.e_1}{e_2'}}
-
 \inferrule[SS-E-Check-Ok]
@@
 {\rcheck{r}{\strin{s}}{x.e_1}{e_2} \mapsto [\rstr{s} / x]e_1}
+
 \inferrule[SS-E-Check-NotOk]
@@
 \inferrule[P-E-Replace]
-{\iota_1 \treduces \rx{r} \\ \iota_2 \treduces \tstr{s_2} \\ \iota_3 \treduces \tstr{s_3} \\ \lsubst{r}{s_2}{s_3} = s} 
-{\preplace{\iota_1}{\iota_2}{\iota_3} \treduces \tstr{s}}
+{\iota_1 \treduces \rx{r} \\ \iota_2 \treduces \tstr{s_2} \\ \iota_3 \treduces \tstr{s_3} } 
+{\preplace{\iota_1}{\iota_2}{\iota_3} \treduces \tstr{\lsubst{r}{s_2}{s_3} }}
 
@@
 {
-  \iota_r \treduces \rx{r} \\
+  \iota_x \treduces \rx{r} \\
   \iota \treduces \tstr{s} \\ 
@@
 {
-  \tcheck{\iota_r}{\iota}{\iota_1}{\iota_2} \treduces \dot{v_1}
+  \tcheck{\iota_x}{\iota}{\iota_1}{\iota_2} \treduces \dot{v_1}
 }
@@
 \inferrule[P-E-Check-NotOK]
-{\iota_r \treduces \rx{r} \\ \iota \treduces \tstr{s} \\ s \not \in \lang{r} \\ \iota_2 \treduces \dot{v_2}}
-{\tcheck{\iota_r}{\iota}{\iota_1}{\iota_2} \treduces \dot{v_2}}
+{\iota_x \treduces \rx{r} \\ \iota \treduces \tstr{s} \\ s \not \in \lang{r} \\ \iota_2 \treduces \dot{v_2}}
+{\tcheck{\iota_x}{\iota}{\iota_1}{\iota_2} \treduces \dot{v_2}}
 \end{mathpar}

### 2014-12-14 sanitation-psp14/tr/strings_chapter.tex
prog.

@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+% LAMBDAP SECTION
 
@@
 
-\begin{thm}[Safety for $\lambdap$] If $\emptyset \vdash \iota : \tau$ 
-  then $\iota \sreduces \dot v$ and $\emptyset \vdash \dot v : \tau$.
-\end{thm}
-
-We can also define canonical forms for regular expressions and strings in the usual way:
+This section follows the same structure as the safety proof for $\lambdas$ --
+we prove type safety for a small-step semantics, prove a semantic correspondence,
+and then transfer the safety result to the big-step semantics in the paper.
+
+Our proofs here are more sketchy than in the $\lambdas$ section both because the
+proof method is already demonstrated, and becaues $\lambdap$ is a fairly straightforward
+extension of $\lambda$ for which these results are already estbalished.
 
@@
 \end{lem}
+
+\begin{thm}[Progress]
+If $\emptyset\vdash \iota : \tau$ either $\iota = \dot{v}$ or $\iota \mapsto \iota'$ for some $\iota'$.
+\end{thm}
+\begin{proof}
+The proof proceeds by induction on the typing assumption. We elide contexts where context has obviously not changed.
+
+Consider only the string and regex (non-$\lambda$) fragments of $\lambdap$.
+
+Regular expressions are always values.
+
+For strings, the concatenation cases are trivial.
+
+\begin{itemize}[label=$ $,itemsep=1ex]
+
+\item \textbf{P-T-Case}.
+Suppose $\emptyset \vdash \pstrcase{\iota_1}{\iota_2}{x,y.\iota_3}$.
+By inversion, $\iota_1 : \str$ and so either $\iota_1 \mapsto \iota_1'$
+or by canonical forms, $\iota_1 = \tstr{s_1}$.
+Similarly, $\iota_2 \mapsto \iota_2'$ or else $\iota_2 = \tstr{s_2}$.
+In the former cases, progress occurs via the compatibility rules.
+in the case where both are string values, progress occurs via the case concatenation rule.
+
+\item \textbf{P-T-Replace}.
+Suppose $\emptyset \vdash \preplace{\iota_1}{\iota_2}{\iota_3}$.
+By inversion, $\iota_1 : \regex$ and so by canonical forms $\iota_1 = \rx{r}$.
+By inversion, $\iota_2 : \str$ and so by induction either $\iota_2 \mapsto \iota_2'$
+or else $\iota_2 = \tstr{s_2}$ for some string $s_2$.
+Similarly, either $\iota_3$ steps or else $\iota_3 = \tstr{s_3}$.
+In case any steps occur, progress occurs. In the remaining case,
+PP-E-Replace applies and so progress occurs.
+
+\item \textbf{P-T-Check}.
+Finally, suppose $\emptyset \vdash \check{\iota_x}{\iota_1}{\iota_2}{\iota_3}$.
+In case any of these step, then progress occurs.
+In the remaining cases, applications of inversion and canonical forms for each
+$\iota_x$ and $\iota_1$ implies that the term at hand equals $\check{\rx{r}}{\tstr{s}}{\iota_2}{\iota_3}$,
+which evaluates to either $\iota_2$ or $\iota_3$.
+
+\end{itemize}
+\end{proof}
+
+\begin{thm}[Preservation]
+If $\emptyset \vdash \iota : \tau$ and $\iota \mapsto \iota'$ then $\emptyset \vdash \iota' : \tau$.
+\end{thm}
+\begin{proof}
+\todo{Fill in if there's time. Otherwise there's nothing interesting going on.}
+\end{proof}
+
+\begin{thm}[Safety for Small-Step Semantics]
+If $\iota \sreduces \dot{v}$ and $\iota \mapsto \iota'$ then $\iota' \sreduces \dot{v}$. 
+\end{thm}
+\begin{proof}
+A direct result from progress and preservation.
+\end{proof}
+
+We only prove semantic correspondence in one direction; again, whereas $\lambdas$ proofs
+were detailed, here we provide a less verbosy proof.
+
+\begin{thm}[Semantic Correspondence]
+If $\iots \mapsto^* \iota'$ then $\iota \sreduces \iota'$.
+\end{thm}
+\begin{proof}
+By induction on the structure of $\iota$.
+
+\begin{itemize}[label=$ $,itemsep=1ex]
+\end{itemize}
+\end{proof}
+
+\begin{thm}[Safety for $\lambdap$] If $\emptyset \vdash \iota : \tau$ 
+  then $\iota \sreduces \dot v$ and $\emptyset \vdash \dot v : \tau$.
+\end{thm}
+\begin{proof}
+If $\emptyset \vdash \iota : \tau$ then $\iota \mapsto^* \iota'$.
+Therefore, $\iota \sreduces \iota'$ by part 2 of the semantic correspondence theorem.
+
+Since $\emptyset \vdash \iota : \tau$ and $\iota \mapsto^* \iota'$, it follows that $\emptyset \vdash \iota' : \tau$ by type safety for the small step semantics.
+
+\end{proof}
+
+We can also define canonical forms for regular expressions and strings in the usual way:
 
@@
 \begin{mathpar}
-\inferrule[SP-E-Abs]
-{ \ }
-{\lambda x.e \sreduces \lambda x.e}
-
-\inferrule[SP-E-App]
-{ \iota_1 \sreduces \lambda x . \iota_3 \\  \iota_2 \sreduces \dot{v}_2 \\ [\dot{v}_2 / x] \iota_3 \sreduces \dot{v}_3}
-{ \iota_1(\iota_2) \sreduces \dot{v}_3}
-
-\inferrule[SP-E-Str]
-{ \ }
-{\tstr{s} \treduces \tstr{s}}
-
-\inferrule[SP-E-Rx]
-{ \ }
-{\rx{r} \treduces \rx{r}}
-
-\inferrule[SP-E-Concat]
-{\iota_1 \treduces \tstr{s_1} \\ \iota_2 \treduces \tstr{s_2}} 
-{\tconcat{\iota_1}{\iota_2} \treduces \tstr{s_1 s_2}} % ???
-
-\inferrule[SP-E-Case-$\epsilon$]
-{
-  \iota_1 \treduces \tstr{ \epsilon } \\
-  \iota_2 \treduces \dot{v_2} \\
-}
-{
-  \pstrcase{\iota_1}{\iota_2}{x,y.\iota_3} \treduces \dot{v_2}
-}
-
-\inferrule[SP-E-Case-Concat]
-{
-  \iota_1 \treduces \tstr{as} \\
-  [\tstr{a},\tstr{s}/x,y] \iota_3 \treduces \dot{v} 
-}
-{
-  \pstrcase{\iota_1}{\iota_2}{x,y.\iota_3} \treduces \dot{v}
-}
-
-\inferrule[SP-E-Replace]
-{\iota_1 \treduces \rx{r} \\ \iota_2 \treduces \tstr{s_2} \\ \iota_3 \treduces \tstr{s_3} \\ \lsubst{r}{s_2}{s_3} = s} 
-{\preplace{\iota_1}{\iota_2}{\iota_3} \treduces \tstr{s}}
-
-\inferrule[SP-E-Check-OK]
-{
-  \iota_r \treduces \rx{r} \\
-  \iota \treduces \tstr{s} \\ 
-s \in \lang{r} \\ 
-\iota_1 \treduces \dot{v_1} 
-}
-{
-  \tcheck{\iota_r}{\iota}{\iota_1}{\iota_2} \treduces \dot{v_1}
-}
-
-\inferrule[SP-E-Check-NotOK]
-{\iota_r \treduces \rx{r} \\ \iota \treduces \tstr{s} \\ s \not \in \lang{r} \\ \iota_2 \treduces \dot{v_2}}
-{\tcheck{\iota_r}{\iota}{\iota_1}{\iota_2} \treduces \dot{v_2}}
+\inferrule[PP-E-ConcatLeft]
+{\iota_1 \mapsto \iota_1'}
+{\tconcat{\iota_1}{\iota_2} \treduces \tconcat{\iota_1'}{\iota_2}}
+
+\inferrule[PP-E-ConcatRight]
+{\iota_2 \mapsto \iota_2'}
+{\tconcat{\iota_1}{\iota_2} \treduces \tconcat{\iota_1}{\iota_2'}}
+
+\inferrule[PP-E-Concat]
+{\ }
+{\tconcat{\tstr{s_1}}{\tstr{s_2}} \treduces \tstr{s_1s_2}}
+
+\inferrule[PP-E-CaseLeft]
+{ \iota_1 \mapsto \iota_1' }
+{ \pstrcase{\iota_1}{\iota_2}{x,y.\iota_3} \mapsto \pstrcase{\iota_1'}{\iota_2}{x,y.\iota_3} }
+
+\inferrule[PP-E-CaseRight]
+{ \iota_2 \mapsto \iota_2' }
+{ \pstrcase{\dot{v}}{\iota_2}{x,y.\iota_3} \mapsto \pstrcase{\dot{v}}{\iota_2'}{x,y.\iota_3} }
+
+\inferrule[PP-E-Case]
+{ }
+{ \pstrcase{\tstr{s_1}}{\tstr{s_2}}{x,y.\iota_3} \mapsto \tstr{s_1s_2}}
+
+\inferrule[PP-E-ReplaceLeft]
+{\iota_1 \mapsto \iota_1'}
+{ \preplace{\iota_1}{\iota_2}{\iota_3} \mapsto \preplace{\iota_1'}{\iota_2}{\iota_3} }
+
+\inferrule[PP-E-ReplaceMid]
+{\iota_2 \mapsto \iota_2'}
+{ \preplace{\rx{r}}{\iota_2}{\iota_3} \mapsto \preplace{\rx{r}}{\iota_2'}{\iota_3} }
+
+\inferrule[PP-E-ReplaceRight]
+{\iota_3 \mapsto \iota_3'}
+{ \preplace{\rx{r}}{\tstr{s_2}}{\iota_3} \mapsto \preplace{\rx{r}}{\tstr{s_2}}{\iota_3'} }
+
+\inferrule[PP-E-Replace]
+{\ }
+{ \preplace{\rx{r}}{\tstr{s_2}}{\tstr{s_3}} \mapsto \tstr{\lsubst{r}{s_2}{s_3}} }
+
+\inferrule[PP-E-Check-Left]
+{ \iota_x \mapsto \iota_x' }
+{ \check{\iota_x}{\iota}{\iota_1}{\iota_2} \mapsto \check{\iota_x}{\iota}{\iota_1}{\iota_2} }
+
+\inferrule[PP-E-Check-Right]
+{ \iota \mapsto \iota' }
+{ \check{\rx{r}}{\iota}{\iota_1}{\iota_2} \mapsto \check{\rx{r}}{\iota'}{\iota_1}{\iota_2} }
+
+\inferrule[PP-E-Check-OK]
+{ s \in \lang{r} }
+{ \check{\rx{r}}{\tstr{s}}{\iota_1}{\iota_2} \mapsto \iota_1}
+
+\inferrule[PP-E-Check-OK]
+{ s \not\in \lang{r} }
+{ \check{\rx{r}}{\tstr{s}}{\iota_1}{\iota_2} \mapsto \iota_2}
+
 \end{mathpar}
-\caption{Small step semantics for $\lambdap$}.
+\caption{Small step semantics for $\lambdap$ (extends L-E rules)}.
 \label{fig:sdlambdap}

### 2014-12-14 sanitation-psp14/tr/strings_chapter.tex
Began texing plane work.

@@
 \begin{proof}
-By induction on the structure of e.
+By induction on the structure of $e$.
+
 \begin{itemize}[label=$ $,itemsep=1ex]
@@
 By induction, $e = (\lambda x . e')v_2 \sreduces v$.
+
+\item \todo{Other lambda cases}
+
+\item \textbf{$s = \concat{e_1}{e_2}$}.
+The only big-step rule that applies is S-E-Concat,
+so $e \sreduces v$ where $v = \rstr{s_1s_2}$,
+$e_1 \sreduces \rstr{s_1}$, and $e_2 \sreduces \rstr{s_2}$.
+
+If $e_1 = \rstr{e_1}$ and $e_2 = \rstr{e_2}$ then $e \mapsto v$.
+Otherwise either $e_1 \mapsto e_1'$ or else $e_2 \mapsto e_2'$.
+
+In the former case, $e \mapsto \concat{e_1'}{e_2}$ by S-E-Concat-Left.
+By induction, $e_1' \sreduces \rstr{s_1}$ and so 
+$\concat{e_1'}{e_2} \sreduces v$.
+
+In the latter case, $e \mapsto \concat{e_1}{e_2'}$ by S-E-Concat-Right.
+By induction, $e_2 \sreduces e_2'$ form which it follows by S-E-Concat
+that $\concat{e_1}{e_2'} \sreduces v$.
+
+\item \textbf{$s = \rstrcase{e_1}{e_2}{x,y.e_3}$}.
+\todo{finish...}
+
 \end{itemize}
@@
 \begin{thm}[Semantic Correspondence]
-If $\iots \mapsto^* \iota'$ then $\iota \sreduces \iota'$.
+If $\iota \mapsto^* \iota'$ then $\iota \sreduces \iota'$.
 \end{thm}
@@
 \begin{itemize}[label=$ $,itemsep=1ex]
+\item \todo{...}
 \end{itemize}

### 2014-12-16 sanitation-psp14/tr/strings_chapter.tex
Updated TR

--- 2014-12-14 sanitation-psp14/tr/strings_chapter.tex
@@
 \newtheorem{ass}[tr]{Assumption}
+\newtheorem{condition}[tr]{Condition}
 
@@
 \abstract{This report contains supporting evidence for claims put forth and explained in the paper ``Statically Typed String Sanitation Inside a Python'' \cite{fulton2014}, including proofs of lemmas
-and theorems asserted in the paper, examples, and additional discussion of the paper's technical content.}
+and theorems asserted in the paper, examples, additional discussion of the paper's technical content, and errata.}
 \begin{document}
 \maketitle
+
+\tableofcontents
+\listoffigures
+\clearpage
+
 \section{Terminology and Notation}
-Theorems and lemmas appearing in \cite{fulton2014} are numbered, while supporting facts appearing only in the Technical Report are lettered.
-Numbered items correspond to the numbering in \cite{fulton2014}.
-
-\section{$\lambdas$}
-
-This section presents proofs of lemmas and theorems about the type systems
-presented in \cite{fulton2014}, the accompanying paper. In addition, we provide some examples to help 
-motivate and explain definitions.
-
-To facilitate the type safety proof, we introduce
-a small step semantics for both $\lambdas$ and $\lambdap$. All theorems in
-this section are proven as stated in \cite{fulton2014}.
-
-\subsection{Head and Tail Operations}
-
-\begin{defn}[Definition of \lhead{r}]\label{def:lhead}
-The relation $\lhead{r}=r'$ is defined in terms of the structure of $r$:
+Theorems and lemmas appearing in \cite{fulton2014} are numbered correspondingly, while supporting facts appearing only in the Technical Report are lettered.
+
+\section{Regular Expressions}
+The syntax of regular expressions over some alphabet $\Sigma$ is shown in Figure \ref{fig:regex}.
+
+\begin{ass}[Regular Expression Congruences]
+We assume regular expressions are implicitly identified up to the following congruences:
 \begin{align*}
-\lhead{r}                 &= \lhead{r, \epsilon} \\
-\lhead{\epsilon, r'}      &= \epsilon \\
-\lhead{a, r'}             &= a \\
-\lhead{r_1\cdot r_2, r'}  &= \lhead{r_1, r_2} \\
-\lhead{r_1 + r_2, r'}     &= \lhead{r_1, r'} + \lhead{r_2, r'} \\
-\lhead{r^*, r'}           &= \lhead{r', \epsilon} + \lhead{r, \epsilon}
+\epsilon\cdot r & \equiv r\\
+r \cdot \epsilon & \equiv r\\
+(r_1\cdot r_2) \cdot r_3 & \equiv r_1 \cdot (r_2 \cdot r_3)\\
+r_1 + r_2 &\equiv r_2 + r_1\\
+(r_1 + r_2) + r_3 &\equiv r_1 + (r_2 + r_3)\\
+\epsilon^* & \equiv \epsilon
 \end{align*}
-\end{defn}
-
-\begin{defn}[Brzozowski's Derivative]\label{def:derivative}
-  The \emph{derivative of $r$ with respect to $s$} is denoted by $\delta_s(r)$
-  and is $\delta_s(r) = \{t | st \in \lang{r}\}$.
-\end{defn}
-
-\begin{defn}[Definition of \ltail{r}]
-The relation $\ltail{r} = r'$ is defined in terms of $\lhead{r}$.
-Note that $\lhead{r} = a_1 + a_2 + ... + a_i$.
-We define $\ltail{r} = \delta_{a_1}(r) + \delta_{a_2}(r) + ... + \delta_{a_i}(r) + \epsilon$.
-\end{defn}
-
-Using these definitions of head and tail, we establish a correctness
-result upon which type soundness for the concatenation operator in
-$\lambdas$ depends.
-
-Throughout the following proofs, we abbreviate $s \in \lang{r}$ by $s \in r$.
-
-\begin{trlem}[Leading characters are in the head] \label{trlemmax}
-If $s = c_1  c_2 \cdots c_m \in \lang{r}$, then
-$c_1 \in \lhead{r}$.
-\end{trlem}
-\begin{proof}
-By structural induction on $r$.
-
-If $r = x$ for some character $x$ then $s \in \lang{r}$ implies $s = x$.
-By definition of lhead, $\lhead{s} = \lhead{s, \epsilon} = \lhead{x, \epsilon} = x$.
-Finally, $x \in \lang{x}$.
-
-If $r = r_1 \cdot r_2$ then $s \in \lang{r}$ implies $s = c_1 s_1 s_2$
-such that $c_1 s_1 \in \lang{r_1}$. 
-By definition of lhead, $\lhead{r_1 \cdot r_2} = \lhead{r_1, r_2} =
-\cdots + \lhead{r_1, \epsilon} = \cdots + \lhead{r_1}$. Therefore, $r_1 \subseteq \lhead{r_1, r_2}$
-and so it suffices to show that $s = c_1 s_1 s_2$ such that $c_1 \in \lhead{r_1}$,
-\todo{this is why we break out the $c_1$ from the $s_1$, both here and going forward}
-which holds by induction because $c_1 s_1 \in \lang{r_1}$.
-
-If $r = r_1 + r_2$ then $s \in r$ implies either $s \in r_1$ or $s \in r_2$.
-Without loss of generality suppose $s \in r_1$.
-Then by induction $s = c_1 \cdots c_m$ such that $c_1 \in \lhead{r_1}$.
-By definition, $\lhead{r} = \lhead{r_1, \epsilon} + \lhead{r_2, \epsilon} = \lhead{r_1} + \cdots$.
-Therefore, $c_1 \in \lhead{r}$.
-
-If $r = r_1^*$ then $s \in r$ implies that $s = s_1 \cdots s_k$ where $k \ge 0$ and $s_i \in \lang{r}$.
-By definition of lhead, $\lhead{r_1^*} = \epsilon + \lhead{r_1}$.
-If $s = \epsilon$, then $s \in \lang{\epsilon} \subseteq \lang{\lhead{r_1}}$.
-If $s = s_1 \cdots s_k$, then $s_1 \in \lang{r}$ and so by induction $s_1 = c_1 \cdots c_k$ such that $c_1 \in \lang{\lhead{r}}$.
-\end{proof}
-
-\begin{trthm}[Correctness of \textsf{ltail}]
-If $s \in \lang{r}$ then $s \in \lang{\lhead{r} \cdot \ltail{r}}$.
-\end{trthm}
-\begin{proof}
-The proof proceeds by structural induction on $r$.
-In each case, we identify sublanguages of the head and tail of $r$ such that 
-$s$ is in the concatenation of the the sublanguages.
-Throughout the proof, we use the fact that $\lang{r'} \subseteq \lang{r}$ and $\lang{q'} \subseteq \lang{q}$
-implies $\lang{r' \cdot q'} \subseteq \lang{r \cdot q}$.
-
-\begin{itemize}[label=$ $,itemsep=1ex]
-
-\item \textbf{Case $r = c$ for $c \in Sigma$}.
-By definition, if $s \in \lang{c}$ then $s = c$.
-By definition of lhead and ltail, $\lhead{c} = \lhead{c, \epsilon} = c$
-and $\lang{\epsilon} \subseteq \lang{\ltail{r}}$.
-Therefore, $c \in \lang{c \cdot \epsilon} \subseteq \lang{\lhead{r} + \ltail{r}}$.
-
-\item \textbf{Case $r = r_1 \cdot r_2$}.
-Suppose $s \in \lang{r}$.
-We may decompose $s$ as $s_1  \cdots s_n$ such that $s_1 = c_1 s_1' \in \lang{r_1}$.
-Note that $c_1 \in \lhead{r_1}$ by \ref{trlemmax}.
-Therefore, $c_1 \in \lhead{r_1 \cdot r_2}$ by definition of lhead.
-It suffices to show that $s_1' s_2 \in \lang{\delta_{c_1}(r)}$.
-Since $s = c_1 s_1' s_2\in \lang{r}$, it follows by the definition of derivative that $s_1' s_2 \in \delta_{c_1}(r)$.
-Therefore, $s = c_1 s_1' s_2 \in \lang{ c_1 \cdot \delta_{c_1} } \subseteq \lang{ \lhead{r} \cdot \ltail{r}}$.
-
-\item \textbf{Case $r = r_1 + r_2$}.
-Suppose $s \in \lang{r_1 + r_2}$
-so that $s \in \lang{r_1}$ or $s \in \lang{r_2}$.
-Consider without loss of generality the case where $s \in \lang{r_1}$.
-By induction, $s \in \lang{\lhead{r_1} \cdot \ltail{r_1}}$
-Note that $\lhead{r_1} \subseteq \lhead{r}$
-and $\ltail{r_1} \subseteq \ltail{r}$.
-Therefore, $\lhead{r_1} \cdot \ltail{r_2} \subseteq \lhead{r_1} \cdot \ltail{r_1}$.
-
-\item \textbf{Case $r = q^*$}.
-Either $s$ is the empty string or else the $k^{th}$ unwinding (i.e., $s = s_1 \cdots s_k$ where $k > 0$ and $s_i \in \lang{q}$).
-
-Suppose $s = \epsilon$.
-By definition of lhead, $\lhead{r} = \lhead{q^*} = \lhead{q^*, \epsilon} = \lhead{\epsilon, \epsilon} + \cdots = \epsilon + \cdots$.
-By definition of ltail, $\ltail{r} = \epsilon \cdots$.
-From these facts we conclude that $s = \epsilon \in \lang{\epsilon \cdot \epsilon} \subseteq \lang{\lhead{r} \cdot \ltail{r}}$.
-
-Suppose instead that $s = s_1 \cdots s_k$ for $k < 0$ where $s_i \in \lang{q}$.
-If $q = \epsilon$ then $s = \epsilon$ and the conclusion follows for the same reason as the epsilon subcase of the character case.
-Otherwise, $s$ is not the empty string and so $s = c_1 s_1 s_2$ where $s_1$ and $s_2$ might be empty.
-By the lemma above, $c_1 \in \lhead{s}$.
-By the definition of derivative, $s_1s_2 \in \delta_{c_1}(s)$.
-Therefore, $s \in \lang{\lhead{s} \cdot \delta_{c_1}(s)} \subseteq \lang{\lhead{r} \cdot \ltail{r}}$, where the set inclusion follows by the definition
-of lhead and ltail.
-\end{itemize}
-\end{proof}
-
-\begin{trexam}[Decomposition is not Reversible]\label{exam:headsandtails}
-$r \not = \lhead{r} \cdot \ltail{r}$.
-\end{trexam}
-\begin{proof}
-A simple counter-example is $ab + cd$. 
-Note that $\lhead{a\cdot b + c\cdot d} = a+c$ and $\ltail{a\cdot b + c\cdot d} = b + d$. 
-Therefore, $\{ad,bc\} \subset \lang{\lhead{a\cdot b + cd\cdot } \cdot \ltail{a\cdot b + c\cdot d}}$ even though neither of these is in $\lang{r}$.
-\end{proof}
-
-Example~\ref{exam:headsandtails} does not imply a counter-example to type soundness because $s \in \lang{r} \implies s \in \lhead{r}\cdot\ltail{r}$ is the
-property required for soundness.
-
-This is a general pattern in string operations: $\lambdas$ simulates -- within the type system -- common operations on strings. 
-If there is an operation for concatenating to strings, we define an operation for concatenating two regular expressions.
-If there is an operation for peeling off the first (n) characters of a string, then we define an operation for 
-converting a regular expression $r$ into a regular expression $r'$ which recognizes any $n^{th}$ suffix of a string in $r$.
-
-It is important to note, however, that the type system need not \emph{exactly} simulate the action of string operations.
-In the case of concatenation, we lose some information because more string values are possible -- according the types --
-than are actually possible in the dynamic semantics.
-Soundness is not lost because the types are conservative in their approximation.
-
-In the case of string replacement, there are \emph{trivial} definitions of substitution (on strings) and replacement (on languages) which over-approximate 
-the effect of a substitution. Closing these gaps in approximation is important, and motivates the string operations portion of this technical report.
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-% Substitution.
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-\subsection{Some Corollaries About Substitution and Language Replacement}
-
-\begin{defn}[$\tt{subst}$]
-We consider several choices in the string operations section.
-\end{defn}
-
-\begin{defn}[$\tt{lreplace}$]
-We consider several choices in the string operations section.
-\end{defn}
-
-\begin{prop}[Closure] \label{thm:total}
-  If $\lang{r}, \lang{r_1}$ and $\lang{r_2}$ are regular languages, then $\lang{\lreplace{r}{r_1}{r_2}}$ is also a regular language.
-\end{prop}
-\begin{proof}
-This result is proven for various formulations in the next section.
-\end{proof}
-
-\begin{prop}[Substitution Correspondence] \label{thm:substcorrespondence}
-  If $s_1 \in \lang{r_1}$ and $s_2 \in \lang{r_2}$ then $\lsubst{r}{s_1}{s_2} \in \lang{\lreplace{r}{r_1}{r_2}}$.
-\end{prop}
-\begin{proof}
-This is exactly the correctness result proven for some pairs of ${\tt subst}$ and ${\tt replace}$ in the previous section.
-\end{proof}
-
-\begin{ass}[Properties of Regular Languages.] \label{thm:regexprops}
-~
+\end{ass}
+
+\begin{ass}[Properties of Regular Languages] \label{thm:regexprops}
+We assume the following properties:
 \begin{enumerate}
@@
 
-Lemma~\ref{thm:regexprops} states some well-known properties about regular expressions.
-
+\section{$\lambdas$}
+The syntax of $\lambdas$ is specified in Figure \ref{fig:glambdas}. The static semantics is specified in Figure \ref{fig:slambdas}.
+
+\subsection{Head and Tail Operations}
+The following correctness conditions must hold for any definition of $\lhead{r}$ and $\ltail{r}$.
+
+\begin{condition}[Correctness of Head] \label{trlemmax}
+If $c_1  s' \in \lang{r}$, then
+$c_1 \in \lang{\lhead{r}}$.
+\end{condition}
+
+\begin{condition}[Correctness of Tail] \label{trlemma:tail-correctness}
+If $c_1 s' \in \lang{r}$ then $s' \in \lang{\ltail{r}}$.
+\end{condition}
+
+For example, we conjecture (but do not here prove) that the definitions below satisfy these conditions. Note that these are slightly amended relative to the published paper.
+
+\begin{defn}[Definition of $\lhead{r}$]\label{def:lhead}
+We first define an auxiliary relation that determines the set of characters that the head might be, tracking the remainder of any sequences that appear:
+\begin{align*}
+\lhead{\epsilon, \epsilon} &= \emptyset\\
+\lhead{\epsilon, r'}      &= \lhead{r', \epsilon} \\
+\lhead{a, r'}             &= \{a\} \\
+\lhead{r_1\cdot r_2, r'}  &= \lhead{r_1, r_2 \cdot r'} \\
+\lhead{r_1 + r_2, r'}     &= \lhead{r_1, r'} \cup \lhead{r_2, r'} \\
+\lhead{r^*, r'}           &= \lhead{r, \epsilon} \cup \lhead{r', \epsilon}
+\end{align*}
+We define $\lhead{r} = a_1 + a_2 + ... + a_i$ iff $\lhead{r, \epsilon} = \{a_1, a_2, ..., a_i\}$.
+\end{defn}
+
+% \begin{trlem}[Correctness of Head] \label{trlemmax}
+% If $c_1  s' \in \lang{r}$, then
+% $c_1 \in {\lhead{r,\epsilon}}$.
+% \end{trlem}
+% \begin{proof}
+% By structural induction on $r$.
+
+% The case for $r = \epsilon$ does not apply because $s$ is non-empty.
+
+% If $r = a$ for some character $a$ then $s \in \lang{r}$ implies $c_1 = a$.
+% By definition $\lhead{a, \epsilon} = \{a\}$.
+% Finally, $a \in \{a\}$ by definition.
+
+% If $r = r_1 \cdot r_2$ then we proceed by case analysis on $r_1$. 
+% \begin{quote}
+% We need not consider the case where $r_1=\epsilon$ by congruence.
+
+% If $r_1=a$ then $c_1=a$ and $\lhead{a \cdot r_2, \epsilon}=\lhead{a, r_2}=a$ and $a \in \{a\}$ by definition.
+
+% If $r_1=r_{11}\cdot r_{12}$, then we have that $\lhead{r_{11} \cdot r_{12} \cdot r_2, \epsilon}=\lhead{r_{11}, r_{12}\cdot r_2}$.
+By the IH we have that $c_1 \in {\lhead{r_{11}\cdot r_{12}, \epsilon}}$.
+By Lemma \ref{trlemma:head-concatenation-1}, we have that $c_1 \in \lhead{r_{11}\cdot r_{12} \cdot r_2, \epsilon}$.
+
+% If $r_1=r_{11} + r_{12}$, then by definition we have that
+
+% \end{quote}
+
+% If $r = r_1 \cdot r_2$ then $s \in \lang{r}$ implies $s = c_1 s_1 s_2$
+% such that $c_1 s_1 \in \lang{r_1}$. 
+% By definition of lhead, $\lhead{r_1 \cdot r_2} = \lhead{r_1, r_2} =
+% \cdots + \lhead{r_1, \epsilon} = \cdots + \lhead{r_1}$. Therefore, $r_1 \subseteq \lhead{r_1, r_2}$
+% and so it suffices to show that $s = c_1 s_1 s_2$ such that $c_1 \in \lhead{r_1}$,
+% \todo{this is why we break out the $c_1$ from the $s_1$, both here and going forward}
+% which holds by induction because $c_1 s_1 \in \lang{r_1}$.
+
+% If $r = r_1 + r_2$ then $s \in r$ implies either $s \in r_1$ or $s \in r_2$.
+% Without loss of generality suppose $s \in r_1$.
+% Then by induction $s = c_1 \cdots c_m$ such that $c_1 \in \lhead{r_1}$.
+% By definition, $\lhead{r} = \lhead{r_1, \epsilon} + \lhead{r_2, \epsilon} = \lhead{r_1} + \cdots$.
+% Therefore, $c_1 \in \lhead{r}$.
+
+% If $r = r_1^*$ then $s \in r$ implies that $s = s_1 \cdots s_k$ where $k \ge 0$ and $s_i \in \lang{r}$.
+% By definition of lhead, $\lhead{r_1^*} = \epsilon + \lhead{r_1}$.
+% If $s = \epsilon$, then $s \in \lang{\epsilon} \subseteq \lang{\lhead{r_1}}$.
+% If $s = s_1 \cdots s_k$, then $s_1 \in \lang{r}$ and so by induction $s_1 = c_1 \cdots c_k$ such that $c_1 \in \lang{\lhead{r}}$.
+% \end{proof}
+
+\begin{defn}[Brzozowski's Derivative]\label{def:derivative}
+  The \emph{derivative of $r$ with respect to $s$} is denoted by $\delta_s(r)$
+  and is $\delta_s(r) = \{t | st \in \lang{r}\}$.
+\end{defn}
+
+\begin{defn}[Definition of $\ltail{r}$]
+If $\lhead{r, \epsilon} = \{a_1, a_2, ..., a_i\}$, then  
+we define $\ltail{r} = \delta_{a_1}(r) + \delta_{a_2}(r) + ... + \delta_{a_i}(r)$.
+\end{defn}
+
+% Using these definitions of head and tail, we establish a correctness
+% result upon which type soundness for the concatenation operator in
+% $\lambdas$ depends.
+
+% \begin{trthm}[Correctness of \textsf{ltail}]
+% If $s \in \lang{r}$ then $s \in \lang{\lhead{r} \cdot \ltail{r}}$.
+% \end{trthm}
+% \begin{proof}
+% The proof proceeds by structural induction on $r$.
+% In each case, we identify sublanguages of the head and tail of $r$ such that 
+% $s$ is in the concatenation of the the sublanguages.
+% Throughout the proof, we use the fact that $\lang{r'} \subseteq \lang{r}$ and $\lang{q'} \subseteq \lang{q}$
+% implies $\lang{r' \cdot q'} \subseteq \lang{r \cdot q}$.
+
+% \begin{itemize}[label=$ $,itemsep=1ex]
+
+% \item \textbf{Case $r = c$ for $c \in Sigma$}.
+% By definition, if $s \in \lang{c}$ then $s = c$.
+% By definition of lhead and ltail, $\lhead{c} = \lhead{c, \epsilon} = c$
+% and $\lang{\epsilon} \subseteq \lang{\ltail{r}}$.
+% Therefore, $c \in \lang{c \cdot \epsilon} \subseteq \lang{\lhead{r} + \ltail{r}}$.
+
+% \item \textbf{Case $r = r_1 \cdot r_2$}.
+% Suppose $s \in \lang{r}$.
+% We may decompose $s$ as $s_1  \cdots s_n$ such that $s_1 = c_1 s_1' \in \lang{r_1}$.
+% Note that $c_1 \in \lhead{r_1}$ by \ref{trlemmax}.
+% Therefore, $c_1 \in \lhead{r_1 \cdot r_2}$ by definition of lhead.
+% It suffices to show that $s_1' s_2 \in \lang{\delta_{c_1}(r)}$.
+% Since $s = c_1 s_1' s_2\in \lang{r}$, it follows by the definition of derivative that $s_1' s_2 \in \delta_{c_1}(r)$.
+% Therefore, $s = c_1 s_1' s_2 \in \lang{ c_1 \cdot \delta_{c_1} } \subseteq \lang{ \lhead{r} \cdot \ltail{r}}$.
+
+% \item \textbf{Case $r = r_1 + r_2$}.
+% Suppose $s \in \lang{r_1 + r_2}$
+% so that $s \in \lang{r_1}$ or $s \in \lang{r_2}$.
+% Consider without loss of generality the case where $s \in \lang{r_1}$.
+% By induction, $s \in \lang{\lhead{r_1} \cdot \ltail{r_1}}$
+% Note that $\lhead{r_1} \subseteq \lhead{r}$
+% and $\ltail{r_1} \subseteq \ltail{r}$.
+% Therefore, $\lhead{r_1} \cdot \ltail{r_2} \subseteq \lhead{r_1} \cdot \ltail{r_1}$.
+
+% \item \textbf{Case $r = q^*$}.
+% Either $s$ is the empty string or else the $k^{th}$ unwinding (i.e., $s = s_1 \cdots s_k$ where $k > 0$ and $s_i \in \lang{q}$).
+
+% Suppose $s = \epsilon$.
+% By definition of lhead, $\lhead{r} = \lhead{q^*} = \lhead{q^*, \epsilon} = \lhead{\epsilon, \epsilon} + \cdots = \epsilon + \cdots$.
+% By definition of ltail, $\ltail{r} = \epsilon \cdots$.
+% From these facts we conclude that $s = \epsilon \in \lang{\epsilon \cdot \epsilon} \subseteq \lang{\lhead{r} \cdot \ltail{r}}$.
+
+% Suppose instead that $s = s_1 \cdots s_k$ for $k < 0$ where $s_i \in \lang{q}$.
+% If $q = \epsilon$ then $s = \epsilon$ and the conclusion follows for the same reason as the epsilon subcase of the character case.
+% Otherwise, $s$ is not the empty string and so $s = c_1 s_1 s_2$ where $s_1$ and $s_2$ might be empty.
+% By the lemma above, $c_1 \in \lhead{s}$.
+% By the definition of derivative, $s_1s_2 \in \delta_{c_1}(s)$.
+% Therefore, $s \in \lang{\lhead{s} \cdot \delta_{c_1}(s)} \subseteq \lang{\lhead{r} \cdot \ltail{r}}$, where the set inclusion follows by the definition
+% of lhead and ltail.
+% \end{itemize}
+% \end{proof}
+
+% \begin{trexam}[Decomposition is not Reversible]\label{exam:headsandtails}
+% $r \not = \lhead{r} \cdot \ltail{r}$.
+% \end{trexam}
+% \begin{proof}
+% A simple counter-example is $ab + cd$. 
+% Note that $\lhead{a\cdot b + c\cdot d} = a+c$ and $\ltail{a\cdot b + c\cdot d} = b + d$. 
+% Therefore, $\{ad,bc\} \subset \lang{\lhead{a\cdot b + cd\cdot } \cdot \ltail{a\cdot b + c\cdot d}}$ even though neither of these is in $\lang{r}$.
+% \end{proof}
+
+% Example~\ref{exam:headsandtails} does not imply a counter-example to type soundness because $s \in \lang{r} \implies s \in \lhead{r}\cdot\ltail{r}$ is the
+% property required for soundness.
+
+% This is a general pattern in string operations: $\lambdas$ simulates -- within the type system -- common operations on strings. 
+% If there is an operation for concatenating to strings, we define an operation for concatenating two regular expressions.
+% If there is an operation for peeling off the first (n) characters of a string, then we define an operation for 
+% converting a regular expression $r$ into a regular expression $r'$ which recognizes any $n^{th}$ suffix of a string in $r$.
+
+% It is important to note, however, that the type system need not \emph{exactly} simulate the action of string operations.
+% In the case of concatenation, we lose some information because more string values are possible -- according the types --
+% than are actually possible in the dynamic semantics.
+% Soundness is not lost because the types are conservative in their approximation.
+
+% In the case of string replacement, there are \emph{trivial} definitions of substitution (on strings) and replacement (on languages) which over-approximate 
+% the effect of a substitution. Closing these gaps in approximation is important, and motivates the string operations portion of this technical report.
+
+% %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+% % Substitution.
+% %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+\subsection{Replacement}
+The following correctness condition must hold for any definition of $\lreplace{r}{r_1}{r_2}$.
+
+\begin{condition}[Replacement Correctness] \label{thm:substcorrespondence}
+  If $s_1 \in \lang{r_1}$ and $s_2 \in \lang{r_2}$ then $$\lsubst{r}{s_1}{s_2} \in \lang{\lreplace{r}{r_1}{r_2}}$$
+\end{condition}
+
+We do not give a particular definition for $\lreplace{r}{r_1}{r_2}$ here.
 %\begin{lem}
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\subsection{The Small Step Semantics}
-
-To prove type safety and the security theorems for the big step semantics, we
-first prove type safety for a small step semantics in Figure~\ref{fig:sdlambdas}
-and then extend this to the big step semantics in Figure~\ref{fig:dlambdas} by
-proving a correspondence between the semantics.
+\subsection{Small Step Semantics of $\lambdas$}
+Figure~\ref{fig:sdlambdas} specifies a small-step operational semantics for $\lambdas$.
 
@@
 \begin{enumerate}
-\item If $\sigma = \stringin{r}$ then $u = \rstr{s}$.
-\item If $\sigma = \sigma_1 \rightarrow \sigma_2$ then $v = \lambda x . e'$ and $s \in \lang{r}$.
+\item If $\sigma = \stringin{r}$ then $v = \rstr{s}$ and $s \in \lang{r}$.
+\item If $\sigma = \sigma_1 \rightarrow \sigma_2$ then $v = \lambda x . e'$.
 \end{enumerate}
@@
 \begin{proof}
-By inspection of valuation and typing rules.
+By inspection of the static and dynamic semantics.
 \end{proof}
 
-For the sake of completeness, we include a statement of the weaker lemma
-stated in the paper:
-
-\begin{lem}[Canonical Forms for the String Fragment of $\lambdas$]\label{thm:cfs}
-  If $\emptyset \vdash e : \stringin{r}$ and $e \sreduces v$ then $v=\rstr{s}$. 
-\end{lem}
-\begin{proof}
-This fact follows directly from Lemma~\ref{cfgen} and the preservation lemma.
-\end{proof}
-
-\begin{trlem}[Progress of small step semantics.]
-If $\emptyset\vdash e : \sigma$ either $e = v$ or $e \mapsto e'$ for some $e'$.
+\begin{trlem}[Progress]
+If $\emptyset\vdash e : \sigma$ either $e = v$ for some $v$ or $e \mapsto e'$ for some $e'$.
 \end{trlem}
 \begin{proof}
-The proof proceeds by induction on the derivation of $\emptyset\vdash e : \sigma$.
+The proof proceeds by rule induction on the derivation of $\emptyset\vdash e : \sigma$.
 \begin{itemize}[label=$ $,itemsep=1ex]
@@
 \item \textbf{S-T-Stringin-I}. Suppose $\emptyset\vdash \rstr{s} : \stringin{s}$.
-Then $\rstr{s} \ \val$ by SS-E-RStr.
-
-\item \textbf{S-T-Concat}. Suppose $\emptyset\vdash \rconcat{e_1}{e_2} : \stringin{r}$ and $\emptyset e_1 : \stringin{r_1}$ and $\emptyset \vdash e_2 : \stringin{r_2}$.
-By induction, $e_1 \mapsto e_1'$ or $e_1 = v_1$ and similarly
-for $e_2$. 
+Then $e=\rstr{s}$.
+
+\item \textbf{S-T-Concat}. Suppose $\emptyset\vdash \rconcat{e_1}{e_2} : \stringin{r_1 \cdot r_2}$ and $\emptyset \vdash e_1 : \stringin{r_1}$ and $\emptyset \vdash e_2 : \stringin{r_2}$.
+By induction, $e_1 \mapsto e_1'$ or $e_1 = v_1$ and similarly, $e_2 \mapsto e_2'$ or $e_2 = v_2$. 
 If $e_1$ steps, then SS-E-Concat-Left applies and so
@@
 %In the former case, $e \mapsto \strcase{e_1}{e_2'}{x,y.e_3}$.
-By induction and Lemma~\ref{thm:cfs} it follows that $e_1 \mapsto e_1'$ or $e_1 = \rstr{s}$.
+By induction and Canonical Forms it follows that $e_1 \mapsto e_1'$ or $e_1 = \rstr{s}$.
 In the former case, $e$ steps by S-E-Case-Left.
@@
 \item \textbf{S-T-Replace}.
-Suppose $e = \rreplace{r}{e_1}{e_2}$, $\emptyset\vdash e : \stringin{r'}$ and:
+Suppose $e = \rreplace{r}{e_1}{e_2}$, $\emptyset\vdash e : \stringin{\lreplace{r}{r_1}{r_2}}$ and:
 \begin{align}
 &\emptyset\vdash e_1 : \stringin{r_1} \\
-&\emptyset\vdash e_2 : \stringin{r_2} \\
-&\lreplace{r}{r_1}{r_2} = r'
+&\emptyset\vdash e_2 : \stringin{r_2}
 \end{align}
@@
 
-\item \textbf{S-T-SafeCheck}
+\item \textbf{S-T-Check}
 Suppose that $\emptyset\vdash \rcheck{r}{e_0}{x.e_1}{e_2} : \stringin{r}$ and:
 \begin{align}
-&\vdash e_0 : \stringin{r_0}  \\
-x:\stringin{r} &\vdash e_1 : \sigma \\
-&\vdash e_2 : \sigma \label{align6}
+\emptyset &\vdash e_0 : \stringin{r_0}  \\
+\emptyset, x:\stringin{r} &\vdash e_1 : \sigma \\
+\emptyset &\vdash e_2 : \sigma \label{align6}
 \end{align}
@@
 By induction,  $e_0 \mapsto e_0'$ or $e_0 = v$.
-In the former case $e$ steps by SS-E-Check-StepRight.
+In the former case $e$ steps by SS-E-Check-StepLeft.
 Otherwise, $e_0 = \rstr{s}$ by Canonical Forms.
@@
 
-\begin{trlem}[Substitution] \label{lem:subst}
-If $\Psi\ x : \sigma \vdash e : \sigma'$ and $\psi \vdash e' : \sigma$, then
-$\psi \vdash [e' / x] e : \sigma'$.
-\end{trlem}
+\begin{ass}[Substitution] \label{lem:subst}
+If $\Psi, x : \sigma' \vdash e : \sigma$ and $\Psi \vdash e' : \sigma'$, then
+$\Psi \vdash [e' / x] e : \sigma$.
+\end{ass}
 
@@
 \begin{itemize}[label=$ $,itemsep=1ex]
-\item \textbf{$\lambda$ fragment}. Cases SS-T-Var, SS-T-Abs, and SS-T-App are exactly as in a proof of type safety for
+\item \textbf{$\lambda$ fragment}. Cases SS-E-AppLeft, SS-E-AppRight, and SS-E-AppAbs are exactly as in a proof of type safety for
 the simply typed lambda calculus.
@@
 and
-$\emptyset\vdash\rstr{s_2} : \stringin{r_2}$. 
-By Lemma~\ref{lem:cfs}, $s_1 \in \lang{r_1}$ and $s_2 \in  \lang{r_2}$ from which it follows by Lemma~\ref{thm:regexprops}
-that $s_1s_2 \in \lang{r_1r_2}$. Therefore, $\emptyset\vdash \rstr{s_1s_2} : \stringin{r_1r_2}$ by S-T-Rstr.
+$\emptyset\vdash\rstr{s_2} : \stringin{r_2}$ and $\emptyset \vdash \rconcat{\rstr{s_1}}{\rstr{s_2}} : \stringin{r_1 \cdot r_2}$. 
+By Canonical Forms, $s_1 \in \lang{r_1}$ and $s_2 \in  \lang{r_2}$ from which it follows by Lemma~\ref{thm:regexprops}
+that $s_1s_2 \in \lang{r_1\cdot r_2}$. Therefore, $\emptyset\vdash \rstr{s_1s_2} : \stringin{r_1\cdot r_2}$ by S-T-Rstr.
 
@@
 $e \mapsto \strcase{e_1'}{e_2}{x,y.e_3}$
-and $\emptyset \vdash e : \sigma$ and $e_1 \rightarrow e_1'$.
+and $\emptyset \vdash e : \sigma$ and $e_1 \mapsto e_1'$.
 The only rule that applies is S-T-Case, so:
@@
 \emptyset&\vdash e_2 : \sigma \label{align8} \\ 
-x:\stringin{\lhead{r}}, y : \stringin{\ltail{r}} &\vdash e_3 : \sigma \label{align9}
+\emptyset, x:\stringin{\lhead{r}}, y : \stringin{\ltail{r}} &\vdash e_3 : \sigma \label{align9}
 \end{align}
@@
 \item \textbf{SS-E-Case-$\epsilon$-Val}.
-Suppose: \begin{align*}
-&e = \strcase{e_0}{e_2}{x,y.e_3} \\
-&\emptyset\vdash e: \sigma \\
-&e \mapsto e_2
-\end{align*}
-The only rule htat applies is S-T-Case, so $e_2 : \sigma$.
+Suppose $\strcase{e_0}{e_2}{x,y.e_3} \mapsto e_2$. 
+The only rule that applies is S-T-Case, so $\emptyset \vdash e_2 : \sigma$.
 
@@
 and that $\emptyset\vdash e : \sigma$.
-By inversion of S-T-Case:
+The only rule that applies is S-T-Case so:
 \begin{align}
@@
 \emptyset&\vdash e_2 : \sigma \label{align14} \\ 
-x:\stringin{\lhead{r}}, y : \stringin{\ltail{r}} &\vdash e_3 : \sigma \label{align15}
+\emptyset, x:\stringin{\lhead{r}}, y : \stringin{\ltail{r}} &\vdash e_3 : \sigma \label{align15}
 \end{align}
 
-We know that $as \in \lang{r}$ by canonical forms on (\ref{align13}) 
-Therefore, $a \in \lang{\lhead{r}}$ by definition of ${\sf lhead}$.
-Furthermore, $\ltail{r} = ... + \delta_a{r} + ...$ by definition of ${\sf ltail}$.
-Note that $s \in \lang{\delta_a{r}}$ by definition of the derivative, and so
-$s \in \lang{\ltail{r}}$
+We know that $as \in \lang{r}$ by Canonical Forms on (\ref{align13}) 
+Therefore, $a \in \lang{\lhead{r}}$ by Condition \ref{trlemmax} and $s \in \lang{\ltail{r}}$ by Condition \ref{trlemma:tail-correctness}.
 
 From these facts about $a$ and $s$ we know by S-T-Rstr that $\emptyset\vdash \rstr{a} : \stringin{\lhead{r}}$
-and $\emptyset\vdash \rstr{s} : \stringin{\lhead{r}}$.
-It follows by (\ref{lem:subst}) that  $\emptyset\vdash [\rstr{a}, \rstr{s} / x,y] e_3 : \sigma$.
+and $\emptyset\vdash \rstr{s} : \stringin{\ltail{r}}$.
+It follows by Assumption \ref{lem:subst} that  $\emptyset\vdash [\rstr{a}, \rstr{s} / x,y] e_3 : \sigma$.\todo{Cyrus stopped here}
 
@@
 
-\begin{itemize}[label=$ $,itemsep=1ex]
-\end{itemize}
+%\begin{itemize}[label=$ $,itemsep=1ex]
+%\end{itemize}
 \end{proof}
@@
 
-\newpage
-
+\bibliographystyle{abbrv}
+\bibliography{research}
+
+\clearpage
 \renewcommand{\grammarlabel}[2]{#1\hfill#2}
@@
 
-\inferrule[L-E-AppLeft]
+\inferrule[SS-E-AppLeft]
 {e_1 \mapsto e_1'}
@@
 
-\inferrule[L-E-AppRight]
+\inferrule[SS-E-AppRight]
 {e_2 \mapsto e_2'}
@@
 
-\inferrule[L-E-AppAbs]
+\inferrule[SS-E-AppAbs]
 { \ }
@@
 \end{figure}
-
-\bibliographystyle{abbrv}
-\bibliography{research}
 \end{document}

### 2014-12-16 sanitation-psp14/tr/strings_chapter.tex
about to merge.

--- 2014-12-14 sanitation-psp14/tr/strings_chapter.tex
@@
 
-\item \textbf{$s = \rstrcase{e_1}{e_2}{x,y.e_3}$}.
+\item \textbf{$s = \strcase{e_1}{e_2}{x,y.e_3}$}.
 \todo{finish...}
@@
 
+\begin{trlem}[Substitution Lemma] \label{lem:psubst}
+If $\theta, x : \tau \vdash \iota : \tau'$ and $\theta \vdash \iota' : \tau$
+then $\theta \vdash [\iota' / x]\iota : \tau'$.
+\end{trlem}
+
 \begin{thm}[Preservation]
@@
 \begin{proof}
-\todo{Fill in if there's time. Otherwise there's nothing interesting going on.}
+
+The proof proceeds by induction of the derivations of $\emptyset \vdash \iota : \tau$
+and $\iota \mapsto \iota'$.
+
+We treat only the non-lambda fragment.
+
+\begin{itemize}[label=$ $,itemsep=1ex]
+
+\item \textbf{Case SP-E-ConcatLeft}.
+Suppose:
+\begin{align*}
+  \iota &= \concat{\iota_1}{\iota_2} \mapsto \concat{\iota_1'}{\iota_2} \\
+  \emptyset &\vdash \iota : \str \\
+  \iota &\mapsto \iota' 
+\end{align*}
+
+The only appliable typing rule is P-T-Concat, so $\emptyset \vdash \iota_1 : \str$
+and $\emptyset \vdash \iota_2 : \str$ where $\tau = \str$. 
+By induction, $\emptyset \vdash \iota_1' : \str$,
+so $\emptyset \vdash \concat{\iota_1'}{\iota_2} : \str$.
+
+\item \textbf{Case SP-E-ConcatRight}
+\begin{align*}
+  e &= \concat{e_1}{e_2} \mapsto \concat{e_1}{e_2'} \\
+  \emptyset &\vdash e : \str \\
+  \iota &\mapsto \iota' 
+\end{align*}
+
+The only appliable typing rule is P-T-Concat, so $\emptyset \vdash \iota_1 : \str$
+and $\emptyset \vdash \iota_2 : \str$ where $\tau = \str$. 
+By induction, $\emptyset \vdash \iota_1' : \str$,
+so $\emptyset \vdash \concat{\iota_1'}{\iota_2} : \str$.
+
+\end{itemize}
+
 \end{proof}

### 2014-12-16 sanitation-psp14/tr/strings_chapter.tex
merge.

--- 2014-12-16 sanitation-psp14/tr/strings_chapter.tex
@@
 By induction, $e = (\lambda x . e')v_2 \sreduces v$.
+
+\item \todo{Other lambda cases}
+
+\item \textbf{$s = \concat{e_1}{e_2}$}.
+The only big-step rule that applies is S-E-Concat,
+so $e \sreduces v$ where $v = \rstr{s_1s_2}$,
+$e_1 \sreduces \rstr{s_1}$, and $e_2 \sreduces \rstr{s_2}$.
+
+If $e_1 = \rstr{e_1}$ and $e_2 = \rstr{e_2}$ then $e \mapsto v$.
+Otherwise either $e_1 \mapsto e_1'$ or else $e_2 \mapsto e_2'$.
+
+In the former case, $e \mapsto \concat{e_1'}{e_2}$ by S-E-Concat-Left.
+By induction, $e_1' \sreduces \rstr{s_1}$ and so 
+$\concat{e_1'}{e_2} \sreduces v$.
+
+In the latter case, $e \mapsto \concat{e_1}{e_2'}$ by S-E-Concat-Right.
+By induction, $e_2 \sreduces e_2'$ form which it follows by S-E-Concat
+that $\concat{e_1}{e_2'} \sreduces v$.
+
+\item \textbf{$s = \strcase{e_1}{e_2}{x,y.e_3}$}.
+\todo{finish...}
+
 \end{itemize}
@@
 
-%\begin{itemize}[label=$ $,itemsep=1ex]
-%\end{itemize}
+\begin{itemize}[label=$ $,itemsep=1ex]
+\item \todo{...}
+\end{itemize}
 \end{proof}

### 2014-12-16 sanitation-psp14/tr/strings_chapter.tex
fixed some errors.

@@
 
-Our proofs here are more sketchy than in the $\lambdas$ section both because the
-proof method is already demonstrated, and becaues $\lambdap$ is a fairly straightforward
-extension of $\lambda$ for which these results are already estbalished.
-
 \begin{lem}[Canonical Forms for Target Language]
@@
 \begin{proof}
-The proof proceeds by induction on the typing assumption. We elide contexts where context has obviously not changed.
-
+The proof proceeds by induction on the typing assumption. 
 Consider only the string and regex (non-$\lambda$) fragments of $\lambdap$.
-
-Regular expressions are always values.
-
-For strings, the concatenation cases are trivial.
-
 \begin{itemize}[label=$ $,itemsep=1ex]
@@
 
-\item \textbf{Case SP-E-ConcatLeft}.
+\item \textbf{Case PS-E-ConcatLeft}.
 Suppose:
 \begin{align*}
-  \iota &= \concat{\iota_1}{\iota_2} \mapsto \concat{\iota_1'}{\iota_2} \\
+  \iota &= \rconcat{\iota_1}{\iota_2} \mapsto \rconcat{\iota_1'}{\iota_2} \\
   \emptyset &\vdash \iota : \str \\
@@
 
-The only appliable typing rule is P-T-Concat, so $\emptyset \vdash \iota_1 : \str$
+The only applicable typing rule is P-T-Concat, so $\emptyset \vdash \iota_1 : \str$
 and $\emptyset \vdash \iota_2 : \str$ where $\tau = \str$. 
 By induction, $\emptyset \vdash \iota_1' : \str$,
-so $\emptyset \vdash \concat{\iota_1'}{\iota_2} : \str$.
-
-\item \textbf{Case SP-E-ConcatRight}
+so $\emptyset \vdash \rconcat{\iota_1'}{\iota_2} : \str$.
+
+\item \textbf{Case PS-E-ConcatRight}
 \begin{align*}
-  e &= \concat{e_1}{e_2} \mapsto \concat{e_1}{e_2'} \\
+  e &= \rconcat{e_1}{e_2} \mapsto \rconcat{e_1}{e_2'} \\
   \emptyset &\vdash e : \str \\
@@
 
-The only appliable typing rule is P-T-Concat, so $\emptyset \vdash \iota_1 : \str$
+The only applicable typing rule is P-T-Concat, so $\emptyset \vdash \iota_1 : \str$
 and $\emptyset \vdash \iota_2 : \str$ where $\tau = \str$. 
 By induction, $\emptyset \vdash \iota_1' : \str$,
-so $\emptyset \vdash \concat{\iota_1'}{\iota_2} : \str$.
+so $\emptyset \vdash \rconcat{\iota_1}{\iota_2'} : \str$.
+
+\todo{Everything above in this proof was sleep deprived plane work -- need to check}
+
+\item \textbf{Case PS-E-Concat}
+Let $e = \rconcat{\rstr{s_1}}{\rstr{s_2}} \mapsto \rstr{s_1s_2}$. 
+The only rule that applies is P-T-Concat, so $\theta \vdash e : \str$.
+By canonical forms, $\theta \rstr{s_1s_2} : \str$.
+
+\item \textbf{Case PS-E-CaseLeft}
+Let $\iota = \strcase{\iota_1}{\iota_2}{x,y.\iota_3} \mapsto \strcase{\iota_1'}{\iota_2}{x,y.\iota_3}$
+when $\iota_1 \mapsto \iota_1'$.
+The only rule that applies is P-T-Case, so $\theta \vdash \iota : \tau$ where:
+\begin{align*}
+  \theta &\vdash \iota_1 : \str 
+  \theta &\vdash \iota_2 : \tau
+  \theta, x : \str, y : \str &\vdash \iota_3 : \tau
+\end{align*}
+By induction, $\theta \vdash \iota_1' : \str$.
+By P-T-Case, $\thet \vdash \strcase{\iota_1}{\iota_2}{x,y.\iota_3} : \tau$.
+
+\item \textbf{Case PS-E-CaseRight}
+Let $\iota = \strcase{\iota_1}{\iota_2}{x,y.\iota_3} \mapsto \strcase{\iota_1}{\iota_2}{x,y.\iota_3}$
+when $\iota_1 \mapsto \iota_1'$.
+The only rule that applies is P-T-Case, so $\theta \vdash \iota : \tau$ where:
+\begin{align*}
+  \theta &\vdash \iota_1 : \str 
+  \theta &\vdash \iota_2 : \tau
+  \theta, x : \str, y : \str &\vdash \iota_3 : \tau
+\end{align*}
+By induction, $\theta \vdash \iota_1' : \str$.
+By P-T-Case, $\thet \vdash \strcase{\iota_1}{\iota_2}{x,y.\iota_3} : \tau$.
 
@@
 \small
-$\fbox{\inferrule{}{\iota \treduces \dot{v}}}$
+$\fbox{\inferrule{}{\iota \mapsto \iota}}$
 
 \begin{mathpar}
-\inferrule[PP-E-ConcatLeft]
+\inferrule[PS-E-ConcatLeft]
 {\iota_1 \mapsto \iota_1'}
@@
 
-\inferrule[PP-E-ConcatRight]
+\inferrule[PS-E-ConcatRight]
 {\iota_2 \mapsto \iota_2'}
@@
 
-\inferrule[PP-E-Concat]
+\inferrule[PS-E-Concat]
 {\ }
@@
 
-\inferrule[PP-E-CaseLeft]
+\inferrule[PS-E-CaseLeft]
 { \iota_1 \mapsto \iota_1' }
@@
 
-\inferrule[PP-E-CaseRight]
-{ \iota_2 \mapsto \iota_2' }
-{ \pstrcase{\dot{v}}{\iota_2}{x,y.\iota_3} \mapsto \pstrcase{\dot{v}}{\iota_2'}{x,y.\iota_3} }
-
-\inferrule[PP-E-Case]
+\inferrule[PS-E-Case-Epsilon]
 {  }
-{ \pstrcase{\tstr{s_1}}{\tstr{s_2}}{x,y.\iota_3} \mapsto \tstr{s_1s_2}}
-
-\inferrule[PP-E-ReplaceLeft]
+{ \pstrcase{\epsilon}{\iota_2}{x,y.\iota_3} \mapsto \iota_2}
+
+\inferrule[PS-E-Case]
+{ }
+{ \pstrcase{\tstr{as}}{\iota_2}{x,y.\iota_3} \mapsto \tstr{as}}
+
+\inferrule[PS-E-ReplaceLeft]
 {\iota_1 \mapsto \iota_1'}
@@
 
-\inferrule[PP-E-ReplaceMid]
+\inferrule[PS-E-ReplaceMid]
 {\iota_2 \mapsto \iota_2'}
@@
 
-\inferrule[PP-E-ReplaceRight]
+\inferrule[PS-E-ReplaceRight]
 {\iota_3 \mapsto \iota_3'}
@@
 
-\inferrule[PP-E-Replace]
+\inferrule[PS-E-Replace]
 {\ }
@@
 
-\inferrule[PP-E-Check-Left]
+\inferrule[PS-E-Check-Left]
 { \iota_x \mapsto \iota_x' }
@@
 
-\inferrule[PP-E-Check-Right]
+\inferrule[PS-E-Check-Right]
 { \iota \mapsto \iota' }
@@
 
-\inferrule[PP-E-Check-OK]
+\inferrule[PS-E-Check-OK]
 { s \in \lang{r} }
@@
 
-\inferrule[PP-E-Check-OK]
+\inferrule[PS-E-Check-NotOK]
 { s \not\in \lang{r} }

### 2014-12-16 sanitation-psp14/tr/strings_chapter.tex
case and replace for p progress.

@@
 Let $e = \rconcat{\rstr{s_1}}{\rstr{s_2}} \mapsto \rstr{s_1s_2}$. 
-The only rule that applies is P-T-Concat, so $\theta \vdash e : \str$.
-By canonical forms, $\theta \rstr{s_1s_2} : \str$.
+The only rule that applies is P-T-Concat, so $\emptyset \vdash e : \str$.
+By canonical forms, $\emptyset \rstr{s_1s_2} : \str$.
 
@@
 when $\iota_1 \mapsto \iota_1'$.
-The only rule that applies is P-T-Case, so $\theta \vdash \iota : \tau$ where:
+The only rule that applies is P-T-Case, so $\emptyset \vdash \iota : \tau$ where:
 \begin{align*}
-  \theta &\vdash \iota_1 : \str 
-  \theta &\vdash \iota_2 : \tau
-  \theta, x : \str, y : \str &\vdash \iota_3 : \tau
+  \emptyset &\vdash \iota_1 : \str  \\
+  \emptyset &\vdash \iota_2 : \tau \\
+  \emptyset, x : \str, y : \str &\vdash \iota_3 : \tau 
 \end{align*}
-By induction, $\theta \vdash \iota_1' : \str$.
-By P-T-Case, $\thet \vdash \strcase{\iota_1}{\iota_2}{x,y.\iota_3} : \tau$.
-
-\item \textbf{Case PS-E-CaseRight}
-Let $\iota = \strcase{\iota_1}{\iota_2}{x,y.\iota_3} \mapsto \strcase{\iota_1}{\iota_2}{x,y.\iota_3}$
-when $\iota_1 \mapsto \iota_1'$.
-The only rule that applies is P-T-Case, so $\theta \vdash \iota : \tau$ where:
+By induction, $\emptyset \vdash \iota_1' : \str$.
+By P-T-Case, $\emptyset \vdash \strcase{\iota_1}{\iota_2}{x,y.\iota_3} : \tau$.
+
+\item \textbf{Case PS-E-CaseEpsilon}
+Let $\iota = \strcase{\rstr{\epsilon}}{\iota_2}{x,y.\iota_3} \mapsto \iota_2$.
+The only rule that applies is P-T-Case, so $\emptyset \vdash \iota : \tau$ where
+$\iota_2 : \tau$.
+
+\item \textbf{Case PS-E-Case}
+Let $\iota = \strcase{\rstr{as}}{\iota_2}{x,y.\iota_3} \mapsto [\rstr{a}, \rstr{s} / x,y] \iota_3$
+The only rule that applies is P-T-Case, so $\emptyset \vdash \iota : \tau$ where:
 \begin{align*}
-  \theta &\vdash \iota_1 : \str 
-  \theta &\vdash \iota_2 : \tau
-  \theta, x : \str, y : \str &\vdash \iota_3 : \tau
+  \emptyset &\vdash \iota_1 : \str \\
+  \emptyset &\vdash \iota_2 : \tau \\
+  \emptyset, x : \str, y : \str &\vdash \iota_3 : \tau
 \end{align*}
-By induction, $\theta \vdash \iota_1' : \str$.
-By P-T-Case, $\thet \vdash \strcase{\iota_1}{\iota_2}{x,y.\iota_3} : \tau$.
+The result follows by the substitution lemma.
+
+\item \textbf{Case PS-E-ReplaceLeft}
+Let $\iota = \rreplace{\iota_1}{\iota_2}{\iota_3} \mapsto \rreplace{\iota_1'}{\iota_2}{\iota_3}$
+where $\iota_1 \mapsto \iota_1'$.
+The applicable typing rule is P-T-Replace, so $\emptyset \vdash \iota : \str$ where:
+\begin{align*}
+  \emptyset &\vdash \iota_1 : \regex \\
+  \emptyset &\vdash \iota_2 : \str \\
+  \emptyset &\vdash \iota_3 : \str
+\end{align*}
+
+By induction, $\emptyset \vdash \iota_1' : \regex$.
+Therefore, $\emptyset \vdash \rreplace{\iota_1'}{\iota_2}{\iota_3}$.
+
+\item \textbf{Case PS-E-ReplaceMid}
+Let $\iota = \rreplace{\iota_1}{\iota_2}{\iota_3} \mapsto \rreplace{\iota_1}{\iota_2'}{\iota_3}$
+where $\iota_2 \mapsto \iota_2'$.
+The applicable typing rule is P-T-Replace, so $\emptyset \vdash \iota : \str$ where:
+\begin{align*}
+  \emptyset &\vdash \iota_1 : \regex \\
+  \emptyset &\vdash \iota_2 : \str \\
+  \emptyset &\vdash \iota_3 : \str
+\end{align*}
+
+By induction, $\emptyset \vdash \iota_2' : \str$.
+Therefore, $\emptyset \vdash \rreplace{\iota_1}{\iota_2'}{\iota_3}$.
+
+\item \textbf{Case PS-E-ReplaceRight}
+Let $\iota = \rreplace{\iota_1}{\iota_2}{\iota_3} \mapsto \rreplace{\iota_1}{\iota_2}{\iota_3'}$
+where $\iota_3 \mapsto \iota_3'$.
+The applicable typing rule is P-T-Replace, so $\emptyset \vdash \iota : \str$ where:
+\begin{align*}
+  \emptyset &\vdash \iota_1 : \regex \\
+  \emptyset &\vdash \iota_2 : \str \\
+  \emptyset &\vdash \iota_3 : \str
+\end{align*}
+
+By induction, $\emptyset \vdash \iota_3' : \str$.
+Therefore, $\emptyset \vdash \rreplace{\iota_1}{\iota_2}{\iota_3'}$.
+
+\item \textbf{Case PS-E-Replace}
+Let $\iota = \rreplace{\rx{r}}{\rstr{s_2}}{\rstr{s_3}} \mapsto \rstr{\lreplace{r}{s_2}{s_3}}$.
+The applicable typing rule is P-T-Replace, so $\emptyset \vdash \iota : \str$.
+The result follows by canonical forms.
 

### 2014-12-16 sanitation-psp14/tr/strings_chapter.tex
lambdap preservation.

@@
 
-\todo{Everything above in this proof was sleep deprived plane work -- need to check}
+\todo{Everything above in the $\lambdap$ section was sleep deprived plane work -- need to check}
 
@@
 
+\item \textbf{Case PS-E-CheckLeft}
+Let $\iota = \rcheck{\iota_x}{\iota_1}{\iota_2}{\iota_3} \mapsto \rcheck{\iota_x'}{\iota_1}{\iota_2}{\iota_3}$
+where $\iota_x \mapsto \iota_x'$.
+The applicable typing rule is P-T-Check, so $\emptyset \vdash \iota : \tau$ where:
+\begin{align*}
+  \emptyset &\vdash \iota_x : \regex \\
+  \emptyset &\vdash \iota_1 : \str \\
+  \emptyset &\vdash \iota_2 : \tau \\
+  \emptyset &\vdash \iota_3 : \tau
+\end{align*}
+By induction, $\iota_x : \regex$. Therefore,
+$\emptyset \rcheck{\iota_x'}{\iota_1}{\iota_2}{\iota_3} : \tau$.
+
+\item \textbf{Case PS-E-CheckRight}
+Let $\iota = \rcheck{\iota_x}{\iota_1}{\iota_2}{\iota_3} \mapsto \rcheck{\iota_x}{\iota_1'}{\iota_2}{\iota_3}$
+where $\iota_1 \mapsto \iota_1'$.
+The applicable typing rule is P-T-Check, so $\emptyset \vdash \iota : \tau$ where:
+\begin{align*}
+  \emptyset &\vdash \iota_x : \regex \\
+  \emptyset &\vdash \iota_1 : \str \\
+  \emptyset &\vdash \iota_2 : \tau \\
+  \emptyset &\vdash \iota_3 : \tau
+\end{align*}
+By induction, $\iota_1' : \str$. Therefore,
+$\emptyset \rcheck{\iota_x}{\iota_1'}{\iota_2}{\iota_3} : \tau$.
+
+\item \textbf{Case PS-E-Check-Ok} 
+Let $\iota = \rcheck{\rx{r}}{\rstr{s}}{\iota_2}{\iota_3} \mapsto \iota_2$ and $s \in \lang{r}$.
+The applicable typing rule is P-T-Check, so $\emptyset \vdash \iota : \tau$ where
+$\emptyset \vdash \iota_2 : \tau$.
+
+\item \textbf{Case PS-E-Check-NotOk} 
+Let $\iota = \rcheck{\rx{r}}{\rstr{s}}{\iota_2}{\iota_3} \mapsto \iota_3$ where $s \not \in \lang{r}$.
+The applicable typing rule is P-T-Check, so $\emptyset \vdash \iota : \tau$ where
+$\emptyset \vdash \iota_3 : \tau$.
+
 \end{itemize}
@@
 \inferrule[P-T-Check]
-{\tctx \vdash \iota_r : \regex \\ \tctx \vdash \iota_1 : \str \\ \tctx \vdash \iota_2 : \sigma \\ \tctx \vdash \iota_3 : \sigma}
-{\tctx \vdash \tcheck{\iota_r}{\iota_1}{\iota_2}{\iota_3} : \sigma}
+{\tctx \vdash \iota_x : \regex \\ \tctx \vdash \iota_1 : \str \\ \tctx \vdash \iota_2 : \tau \\ \tctx \vdash \iota_3 : \tau}
+{\tctx \vdash \tcheck{\iota_x}{\iota_1}{\iota_2}{\iota_3} : \tau}
 \end{mathpar}
@@
 
-\inferrule[PS-E-Check-Left]
+\inferrule[PS-E-CheckLeft]
 { \iota_x \mapsto \iota_x' }
-{ \check{\iota_x}{\iota}{\iota_1}{\iota_2} \mapsto \check{\iota_x}{\iota}{\iota_1}{\iota_2} }
-
-\inferrule[PS-E-Check-Right]
+{ \rcheck{\iota_x}{\iota}{\iota_1}{\iota_2} \mapsto \rcheck{\iota_x}{\iota}{\iota_1}{\iota_2} }
+
+\inferrule[PS-E-CheckRight]
 { \iota \mapsto \iota' }
-{ \check{\rx{r}}{\iota}{\iota_1}{\iota_2} \mapsto \check{\rx{r}}{\iota'}{\iota_1}{\iota_2} }
+{ \rcheck{\rx{r}}{\iota}{\iota_1}{\iota_2} \mapsto \rcheck{\rx{r}}{\iota'}{\iota_1}{\iota_2} }
 
@@
 { s \in \lang{r} }
-{ \check{\rx{r}}{\tstr{s}}{\iota_1}{\iota_2} \mapsto \iota_1}
+{ \rcheck{\rx{r}}{\tstr{s}}{\iota_1}{\iota_2} \mapsto \iota_1}
 
@@
 { s \not\in \lang{r} }
-{ \check{\rx{r}}{\tstr{s}}{\iota_1}{\iota_2} \mapsto \iota_2}
+{ \rcheck{\rx{r}}{\tstr{s}}{\iota_1}{\iota_2} \mapsto \iota_2}
 

### 2014-12-16 sanitation-psp14/tr/strings_chapter.tex
finished preservation.

@@
 The only applicable typing rule is P-T-Concat, so $\emptyset \vdash \iota_1 : \str$
-and $\emptyset \vdash \iota_2 : \str$ where $\tau = \str$. 
+and $\emptyset \vdash \iota_2 : \str$. 
 By induction, $\emptyset \vdash \iota_1' : \str$,
@@
 The only applicable typing rule is P-T-Concat, so $\emptyset \vdash \iota_1 : \str$
-and $\emptyset \vdash \iota_2 : \str$ where $\tau = \str$. 
+and $\emptyset \vdash \iota_2 : \str$. 
 By induction, $\emptyset \vdash \iota_1' : \str$,
 so $\emptyset \vdash \rconcat{\iota_1}{\iota_2'} : \str$.
-
-\todo{Everything above in the $\lambdap$ section was sleep deprived plane work -- need to check}
 

### 2014-12-16 sanitation-psp14/tr/strings_chapter.tex
some TS work for lambda p and slight cleanup of current translation proof.

@@
 It follows by Assumption \ref{lem:subst} that  $\emptyset\vdash [\rstr{a}, \rstr{s} / x,y] e_3 : \sigma$.\todo{Cyrus stopped here}
+
+\item \textbf{Case SS-E-Replace-Left}.
+Suppose that $e = \rreplace{r}{e_1}{e_2} \mapsto \rreplace{r}{e_1'}{e_2}$ when $e_1 \mapsto e_1'$.
+The only rule that applies is S-T-Replace, so
+$\emptyset \vdash e : \stringin{\lreplace{r}{r_1}{r_2}}$ where:
+\begin{align*}
+\emptyset &\vdash e_1 : \stringin{r_1} \\
+\emptyset &\vdash e_2 : \stringin{r_2}
+\end{align*}
+By induction, $\emptyset \vdash e_1' : \stringin{r_1}$.
+Therefore, $\emptyset \vdash \rreplace{r}{e_1'}{e_2} : \stringin{\lreplace{r}{r_1}{r_2}}$ by S-T-Replace.
+
+\item \textbf{SS-E-Replace-Right}.
+Suppose that $e = \rreplace{r}{e_1}{e_2} \mapsto \rreplace{r}{e_1'}{e_2}$ when $e_1 \mapsto e_1'$.
+The only rule that applies is S-T-Replace, so
+$\emptyset \vdash e : \stringin{\lreplace{r}{r_1}{r_2}}$ where:
+\begin{align*}
+\emptyset &\vdash e_1 : \stringin{r_1} \\
+\emptyset &\vdash e_2 : \stringin{r_2}
+\end{align*}
+By induction, $\emptyset \vdash e_1' : \stringin{r_1}$.
+Therefore, $\emptyset \vdash \rreplace{r}{r_1'}{r_2} : \stringin{\lreplace{r}{r_1}{r_2}}$ by S-T-Replace.
+
+\item \textbf{SS-E-Replace}
+Suppose that $e = \rreplace{r}{\rstr{s_1}}{\rstr{s_2}} \mapsto \rstr{\lsubst{r}{s_1}{s_2}}$ 
+The only rule that applies is S-T-Replace, so
+$\emptyset \vdash e : \stringin{\lreplace{r}{r_1}{r_2}}$ where:
+\begin{align*}
+\emptyset &\vdash \rstr{s_1} : \stringin{r_1} \\
+\emptyset &\vdash \rstr{s_2} : \stringin{r_2}
+\end{align*}
+
+Therefore, it suffices to show that $\emptyset \vdash \rstr{\lsubst{r}{s_1}{s_2}} : \stringin{\lreplace{r}{r_1}{r_2}}$
+or, equivalently, that $\lsubst{r}{s_1}{s_2} \in \lang{\lreplace{r}{r_1}{r_2}}$.
+This follows from the fact that $s_1 \in \lang{r_1}$ and $s_2 \in \lang{r_2}$.
+The only rule that applies is S-T-Replace, so
+$\emptyset \vdash e : \stringin{\lreplace{r}{r_1}{r_2}}$ where:
+\begin{align*}
+\emptyset &\vdash \rstr{s_1} : \stringin{r_1} \\
+\emptyset &\vdash e_2 : \stringin{r_2}
+\end{align*}
+By induction, $\emptyset \vdash e_2' : \stringin{r_2}$.
+Therefore, $\emptyset \vdash \rreplace{r}{\rstr{s_1}}{e_2'} : \stringin{\lreplace{r}{r_1}{r_2}}$ by S-T-Replace.
+
+% bookmark-nrf
+\item
 
@@
 \begin{proof}
-We present a proof by induction on the derivation that $\tctx \vdash e : \sigma$.
-we write $e \leadsto \iota$ as shorthand for the final property.
+We present a proof by induction on the structure of $e$ \todo{why not on the typing derivation?}.
+We write $e \leadsto \iota$ as shorthand for the final property.
 
@@
 \begin{align}
-\trden{\rstr{s}} &= \str{s} \\
+\trden{\rstr{s}} &= \tstr{s} \\
 \trden{\stringin{r}} &= \str \\
@@
 \end{align}
-Note that $\emptyset \vdash \str{s} : \str$ by P-T-Str.
+Note that $\emptyset \vdash \tstr{s} : \str$ by P-T-Str.
 Recall that contexts are standard and, in particular, can be weakened.
@@
 
-Summarily, $\str{s}$ is a term of $\lambdap$ such that $\trden{\tctx} \vdash \str{s} : \trden{\sigma}$
+Summarily, $\tstr{s}$ is a term of $\lambdap$ such that $\trden{\tctx} \vdash \tstr{s} : \trden{\sigma}$
 

### 2014-12-17 sanitation-psp14/tr/strings_chapter.tex
work.

@@
 \newcommand{\preplace}[3]{{\sf replace}(#1;#2;#3)}
+\newcommand{\treplace}[3]{{\sf replace}(#1;#2;#3)}
 \newcommand{\tconcat}[2]{{\sf concat}(#1;#2)} \newcommand{\concat}[2]{{\sf concat}(#1;#2)} % lots of legacy naming around these parts...
@@
 
-\item \textbf{SS-E-Replace}
-Suppose that $e = \rreplace{r}{\rstr{s_1}}{\rstr{s_2}} \mapsto \rstr{\lsubst{r}{s_1}{s_2}}$ 
-The only rule that applies is S-T-Replace, so
-$\emptyset \vdash e : \stringin{\lreplace{r}{r_1}{r_2}}$ where:
-\begin{align*}
-\emptyset &\vdash \rstr{s_1} : \stringin{r_1} \\
-\emptyset &\vdash \rstr{s_2} : \stringin{r_2}
-\end{align*}
-
-Therefore, it suffices to show that $\emptyset \vdash \rstr{\lsubst{r}{s_1}{s_2}} : \stringin{\lreplace{r}{r_1}{r_2}}$
-or, equivalently, that $\lsubst{r}{s_1}{s_2} \in \lang{\lreplace{r}{r_1}{r_2}}$.
-This follows from the fact that $s_1 \in \lang{r_1}$ and $s_2 \in \lang{r_2}$.
-The only rule that applies is S-T-Replace, so
-$\emptyset \vdash e : \stringin{\lreplace{r}{r_1}{r_2}}$ where:
-\begin{align*}
-\emptyset &\vdash \rstr{s_1} : \stringin{r_1} \\
-\emptyset &\vdash e_2 : \stringin{r_2}
-\end{align*}
-By induction, $\emptyset \vdash e_2' : \stringin{r_2}$.
-Therefore, $\emptyset \vdash \rreplace{r}{\rstr{s_1}}{e_2'} : \stringin{\lreplace{r}{r_1}{r_2}}$ by S-T-Replace.
-
-% bookmark-nrf
-\item
-
-\item \textbf{Cases SS-E-Replace-Left, SS-E-Replace-Right, SS-E-Check-StepLeft, SS-E-SafeCoerce-Step, SS-E-Check-StepRight}. \todo{todo}
+%\item \textbf{SS-E-Replace}
+%Suppose that $e = \rreplace{r}{\rstr{s_1}}{\rstr{s_2}} \mapsto \rstr{\lsubst{r}{s_1}{s_2}}$ 
+%The only rule that applies is S-T-Replace, so
+%$\emptyset \vdash e : \stringin{\lreplace{r}{r_1}{r_2}}$ where:
+%\begin{align*}
+%\emptyset &\vdash \rstr{s_1} : \stringin{r_1} \\
+%\emptyset &\vdash \rstr{s_2} : \stringin{r_2}
+%\end{align*}
+%
+%Therefore, it suffices to show that $\emptyset \vdash \rstr{\lsubst{r}{s_1}{s_2}} : \stringin{\lreplace{r}{r_1}{r_2}}$
+%or, equivalently, that $\lsubst{r}{s_1}{s_2} \in \lang{\lreplace{r}{r_1}{r_2}}$.
+%This follows from the fact that $s_1 \in \lang{r_1}$ and $s_2 \in \lang{r_2}$.
+%%The only rule that applies is S-T-Replace, so
+%%$\emptyset \vdash e : \stringin{\lreplace{r}{r_1}{r_2}}$ where:
+%%\begin{align*}
+%%\emptyset &\vdash \rstr{s_1} : \stringin{r_1} \\
+%%\emptyset &\vdash e_2 : \stringin{r_2}
+%%\end{align*}
+%%By induction, $\emptyset \vdash e_2' : \stringin{r_2}$.
+%%Therefore, $\emptyset \vdash \rreplace{r}{\rstr{s_1}}{e_2'} : \stringin{\lreplace{r}{r_1}{r_2}}$ by S-T-Replace.
+%
+%\item \textbf{SS-E-Check-StepLeft}.
+%Suppose that $e = \rcheck{r}{e_0}{x.e_1}{e_2} \mapsto \rcheck{r}{e_0'}{x.e_1}{e_2}$.
+%The only applicable typing rule is S-T-Check. 
+%Therefore, $\emptyset \vdash e : \sigma$ where:
+%\begin{align*}
+%\emptyset &\vdash e_0 : \stringin{r} \\
+%\emptyset, x:\stringin{r} &\vdash e_1 : \sigma \\
+%\emptyset &\vdash e_2 : \sigma 
+%\end{align*}
+%By induction, $\emptyset \vdash e_0' : \stringin{r}$.
+%Therefore, the result follows by P-T-Check.
+%
+%\item \textbf{SS-E-Check-Ok}.
+%Suppose that $e = \rcheck{r}{\rstr{s}}{x.e_1}{e_2} \mapsto \replace{\rstr{s}}{x}{e_1}$
+%where $s \in \lang{r}$. The only applicable typing rule is S-T-Check. 
+%Therefore, $\emptyset \vdash e : \sigma$ where:
+%\begin{align*}
+%\emptyset &\vdash e_0 : \stringin{r} \\
+%\emptyset, x:\stringin{r} &\vdash e_1 : \sigma \\
+%\emptyset &\vdash e_2 : \sigma 
+%\end{align*}
+%By substitution, $\emptyset \vdash \replace{\rstr{s}}{x}{e_1} : \sigma$.
+%
+%\item \textbf{SS-E-Check-NotOk}.
+%Suppose that $e = \rcheck{r}{\rstr{s}}{x.e_1}{e_2} \mapsto e_2$
+%where $s \not \in \lang{r}$. The only applicable typing rule is S-T-Check. 
+%Therefore, $\emptyset \vdash e : \sigma$ where $\emptyset \vdash e_2 : \sigma$.
 
@@
 By induction on the structure of $\iota$.
-
-\begin{itemize}[label=$ $,itemsep=1ex]
-\item \todo{...}
-\end{itemize}
+The proof is similar to the proof for $\lambdas$.
+\todo{if there's time add the proof...}
+
+%\begin{itemize}[label=$ $,itemsep=1ex]
+%\item \todo{...}
+%\end{itemize}
 \end{proof}
@@
 
-We can also define canonical forms for regular expressions and strings in the usual way:
+% We can also define canonical forms for regular expressions and strings in the usual way:
 
@@
 \begin{thm}[Translation Correctness]\label{thm:trcorrect}
-  If $\tctx \vdash e : \sigma$ then 
+  If $\sctx \vdash e : \sigma$ then 
   there exists an $\iota$ such that $\trden{e} = \iota$
-  and $\trden{\tctx} \vdash \iota : \trden{\sigma}$.
-  Furthermore, if $e \sreduces v$ then
+  and $\trden{\sctx} \vdash \iota : \trden{\sigma}$.
+  Furthermore, $e \sreduces v$ and
   $\iota \treduces \dot{v}$ such that
@@
 \begin{proof}
-We present a proof by induction on the structure of $e$ \todo{why not on the typing derivation?}.
+We present a proof by induction on the structure of $e$.
 We write $e \leadsto \iota$ as shorthand for the final property.
@@
 
-\item \textbf{Case $e = \rconcat{e_1}{e_2}$}. This case is an obvious appeal to induction.
+\item \textbf{Case $e = \rconcat{e_1}{e_2}$}. \todo{hand wave lots and lots of symbol pushing.} By induction.
 
@@
 Similarly, $\iota \treduces \dot{v}$ where $[a, t / x, y] \iota_3 \treduces \dot{v}$
+
+\item \textbf{Case $e = \rreplace{r}{e_1}{e_2}$}.
+There is only one applicable typing rule, so suppose
+$\sctx \vdash \rreplace{r}{e_1}{e_2} : \stringin{\lreplace{r}{e_1}{e_2}}$.
+Let $\psi = \trden{\sctx}$.
+Note that $\trden{\rreplace{r}{e_1}{e_2}} = \treplace{\rx{r}}{\iota_1}{\iota_2}$
+when by induction $\trden{e_1} = \iota_1$ and $\trden{e_2} = \iota_2$
+such that $\psi \vdash \iota_1$ and $\psi \vdash \iota_2$.
+It follows by P-T-Replace that $\psi \vdash \treplace{\rx{r}}{\iota_1}{\iota_2} : \str$.
+Finally, note that $\trden{\stringin{\lreplace{r}{e_1}{e_2}}} = \str$.
+
+For evaluation correspondence, note that $\trden{ \rstr{\lreplace{r}{s_1}{s_2}} } = \rstr{ \lreplace{r}{s_1}{s_2} }$ and
+so it suffices to show that $\treplace{\rx{r}}{\iota_1}{\iota_2} \sreduces \rstr{r}{s_1}{s_2}$.
+Note that $\lreplace{r}{e_1}{e_2} \sreduces \rstr{\lreplace{r}{s_1}{s_2}}$ 
+where $e_1 \sreduces \rstr{s_1}$, $e_2 \sreduces \rstr{s_2}$, $r \sreduces r$.
+By induction, $\iota_1 \sreduces \rstr{s_1}, \iota_2 \sreduces \rstr{s_2}$, and $\rx{r} \sreduces \rx{r}$.
+So by S-E-Replace, the sufficient condition holds.
+
+\item \textbf{Case $e = \rcoerce{r}{e'}$}.
+The only applicable tpying rule is S-T-SafeCoerce, so suppose
+$\sctx \vdash \rcoerce{r}{e'} : \stringin{r}$ where
+$\sctx \vdash e' : \stringin{r'}$ and $\lang{r'} \subseteq \lang{r}$.
+By induction, $e' \leadsto \iota$ for some $\iota$.
+Therefore, $\trden{\coerce{r}{e'}} = \iota$ by Tr-SafeCoerce.
+
+For evaluation correspondence, note that $e \sreduces v$ where
+$e' \sreduces v$. The result follows by induction because $e' \leadsto \iota$.
+
+\item \textbf{Case $e = \rcheck{r}{e_1}{x.e_2}{e_3}$}.
+The applicable typing rule is S-T-Check,
+so $\psi \vdash e : \sigma$ where 
+$\psi \vdash e_1 : \stringin{r}$,
+$\psi, x : \stringin{r} \vdash e_2 : \sigma$,
+and $\psi \vdash e_3 : \sigma$.
+By induction and a corresponding substitution principle there exists $\iota_1, \iota_2, \iota_3$ such that
+$e_1 \leadsto \iota_1$, 
+$e_2 \leadsto \iota_2$ in context $\psi, s : \stringin{r}$, and
+$e_3 \leadsto \iota_3$.
+Choose $\iota = \tcheck{\rx{r}}{\iota_1}{\lambda x . \iota_2}{\iota_3}$.
+\todo{hand wave} The result follows by induction.
 
@@
 \inferrule[S-T-Check]
-{\sctx \vdash e_0 : \stringin{r_0} \\ \sctx, x:\stringin{r} \vdash e_1 : \sigma \\ \sctx \vdash e_2 : \sigma}
+{\sctx \vdash e_0 : \stringin{r} \\ \sctx, x:\stringin{r} \vdash e_1 : \sigma \\ \sctx \vdash e_2 : \sigma}
 {\sctx \vdash \rcheck{r}{e_0}{x.e_1}{e_2} : \sigma}
@@
 \inferrule[S-E-Replace]
-{e_1 \sreduces \strin{s_1} \\ e_2 \sreduces \strin{s_2} \\ \lsubst{r}{s_1}{s_2} = s} 
-{\sisubst{r}{e_1}{e_2} \sreduces \sistr{s}}
+{e_1 \sreduces \strin{s_1} \\ e_2 \sreduces \strin{s_2} } 
+{\sisubst{r}{e_1}{e_2} \sreduces \sistr{ \lsubst{r}{s_1}{s_2}  }}
 

### 2015-02-06 thesis-proposal/problem.tex
Updated proposal

--- 2014-09-01 thesis-proposal/problem.tex
@@
 \vspace{-25pt}
-\section{Motivation}\label{motivation}
-\begin{quote}\textit{The recent development of programming languages suggests that the simultaneous achievement of simplicity 
-and generality in language design is a serious unsolved 
-problem.}
--- John Reynolds, 1970 \cite{Reynolds70}\end{quote}
+\section{Introduction}\label{motivation}
+% \begin{quote}\textit{The recent development of programming languages suggests that the simul\-taneous achievement of simplicity 
+% and generality in language design is a serious unsolved 
+% problem.} -- John Reynolds, 1970 \cite{Reynolds70}\end{quote}
 %%One might exp ``generality''
-%Designing a general-purpose programming language along with its supporting tools (collectively, a \emph{programming system}) has long stood as a grand challenge in computing. 
-Achieving both simplicity and generality in a programming language and its associated tools -- collectively, a \emph{programming system} -- requires organizing around a small number of mechanisms that support the realization of many others   as libraries.
-% In such a system, new dialects and revisions of the system would rarely be necessary.
+Well-designed general-purpose programming languages are remarkably stable: they rarely need to be revised or forked into dialects, because programmers can express the constructs that they need just as well in orthogonal libraries.
+%Instead, .
+.
 %Indeed, a preponderence of new dialects is clear evidence against claims of generality.
 %Not all useful abstractions can be  realized, when considered comprehensively, as libraries.% The problem of achieving both simplicity and generality simultaneously with a single system remains unsolved.
-Much progress has been made in the decades since Reynolds' statement above, but a programming system that comprehensively achieves these goals remains elusive, as evidenced by the fact that across all language lineages,  new language and tool {dialects} remain common vehicles for new features and abstractions, both in research and practice.
+Abstraction mechanisms informed by type theory, like  the ML module system \cite{MacQueen:1984:MSM:800055.802036}, have put this ideal within reach. 
+Yet despite their expressive power, new {dialects} of languages like ML continue to arise. Why is this?
+
+% Unlike libraries organized by a module system, language dialects are not trivially composable.%Constructing a software system using libraries written in different dialects is difficult because they cannot always be composed.
+%This suggests that library-based mechanisms are still not general enough to encompass many desirable modes of expression.
 %Library implementations are not yet satisfying.
+%Why would a well-informed programmer consider forming a new dialect? 
+Perhaps the most common justification is that the \emph{syntactic cost} of a desirable construct is not  preserved when it is expressed using general-purpose mechanisms.
+%Syntactic cost is often assessed qualitatively \cite{green1996usability} (though quantitative metrics can also be defined).
+For example, the semantics of regular expressions can be expressed using abstract data types (which in ML are a mode of use of the module system), but as we will detail in Sec.
+\ref{sec:syntax}, this presents some syntactic difficulties.
+For programmers in problem domains  where regular expressions are common, e.g.
+bioinformatics, it is tempting to design a ``domain-specific'' dialect that differs only in that it includes derived syntax (i.e.
+``syntactic sugar'') for regular expression patterns.
+There is no shortage of tools that aid in the construction of such syntactic dialects (e.g.
+Camlp4 \cite{ocaml-manual}, SugarJ \cite{erdweg2011sugarj}, SugarHaskell \cite{erdweg2012layout} and many others).%Put another way, such embeddings may not preserve the cognitive cost associated with a construct (qualitative criteria are generally used to establish this \cite{green1996usability}).
 
-We observe that new system dialects are often motivated by a desire for new concrete syntax, new type system fragments, or new editor services, all {features} of programming systems that library providers usually do not have substantial control over.
-%, encouraged  historically  by the availability of tools like compiler generators and,  more recently, language workbenches \cite{workbenches} and DSL frameworks \cite{dsl}.
+Somewhat more rarely, dialects are motivated by the need to extend a general-purpose language with new derived semantic constructs, i.e.
+constructs that cannot be expressed as purely syntactic sugar, but where the general-purpose language is a suitable translation target.
+As a minimal example, consider System \textbf{F}, i.e.
+the polymorphic lambda calculus (perhaps the simplest general-purpose language) \cite{pfpl}.
+If we wanted to express record types in \textbf{F}, we could not do so simply with derived syntax.
+We could, however, develop a dialect that included the type and term operators relevant to records, specifying their static semantics directly but their dynamic semantics by translation to the polymorphic lambda calculus (using a standard Church-style encoding).\todo{couple more sentences} %Constructs specified by a type-directed translation like this can be considered \emph{derived semantic constructs}.
+
+% express record types as syntactic sugar over the simply-typed lambda calculus with  binary product types.\footnote{Pairs can of course be expressed as syntactic sugar atop records, though one could argue that using binary products as the more primitive concept is simpler.} The static semantics need to be extended with new type and term operators.
+However, the simplest way to express the dynamic semantics of the newly introduced term operators is by translation to nested binary products, so we can leave the operational semantics alone.
+\todo{fill this out} %For example, there are dozens of constructs that go by the name of ``records'' in various languages, each defined by a slightly different collection of primitive operations.
+\todo{examples} %, encouraged  historically  by the availability of tools like compiler generators and,  more recently, language workbenches \cite{workbenches} and DSL frameworks \cite{dsl}.
 Unfortunately, taking this approach makes it substantially more difficult for clients to import high-level abstractions orthogonally.
-Introducing mechanisms that decentralize control over these features could decrease the need for new dialects and thus increase generality.
-But this must done with care, because giving away too much control could weaken important metatheoretic guarantees.
-Moreover, such  mechanisms must come equipped with \emph{modular reasoning principles} for it to be practical for library clients to rely upon them.
-Having to manually resolve conflicts between libraries or reestablish important properties of a program whenever a new library is imported introduces undesirable complexity.
-%It can also make it more difficult to define new sorts of tools for working with the language because they cannot rely on there being a fixed abstract syntax (the canonical \emph{expression problem}).
+
+The problem with designing new language dialects in these situations is that a program written in one dialect cannot safely and idiomatically interface with libraries written in a different dialect.
+%At best, one can hope that the compilers for the two languages target a common intermediate language,
+To do so would require combining the two dialects into a single language, but there is no systematic method to do so, much less one that guarantees that important metatheoretic properties established about each dialect in isolation will be conserved.
+As a simple example, two dialects might be known to have an unambiguous concrete syntax in isolation, but if the syntax is na\"ively combined, ambiguities could easily arise.
+%the ideas housed in one dialect are often only available to programmers willing to do without ideas housed in other dialects.
+%It is thus infeasible to simply allow different contributors to a software system to choose their own favorite dialect for each component they are responsible for.
+%A complex software system written in multiple distinct language dialects is far too unwieldy for it to be viable.
+Without strong modular reasoning principles to rely upon, it is difficult to construct large software systems and ecosystems. 
+%This means that it is difficult to construct large software systems and ecosystems from components written in different dialects. 
+The only reasonable approach for software engineers who must program ``in the large'' is to eschew constructs housed in dialects and settle on a single common language, where they can  rely on the guarantees provided by its type and module system.
+%The designers of these languages may slowly adopt the most generally useful new ideas.
+This has left many useful constructs for programming ``in the small'' obscure.% It is safe to say that in contemporary times, language dialects should be considered harmful.
+% It is little wonder that many programmers eschew esoteric language dialects in favor of languages that stopped evolving decades ago.
+
+\section{Contributions}
+Our aim in this thesis is to make dialect formation less frequently necessary by designing language mechanisms that give library providers the ability to express derived syntactic and semantic constructs from within the language, in a manner that ensures that these constructs can be reasoned about separately and composed arbitrarily.
+
+%Such mechanisms are colloquially known as \emph{extension mechanisms} because the sorts of constructs they are used to express are those that would otherwise require extensions to the language itself.
+%Defining a new construct then corresponds to the usual notion of language extension.
+We will make this intuition more precise later.
+
+%in languages without such mechanisms, expressing such constructs would require extensions to the language itself (a better qualifier might thus be \emph{extension-mitigating}).
+%We will make these notions more precise as we continue.
+%Thus, these mechanisms precisely capture the notions of language extension and language composition.
+
+%The mechanisms we propose are designed to decrease the number of constructs that need to be built in \emph{a priori} by a language designer.
+However, there are still some basic syntactic and semantic decisions that must unavoidably be made.
+
+To provide a coherent vehicle for these mechanisms, we will introduce a new general-purpose language called Wyvern.
+Let us briefly summarize its organization.
+Syntactically, Wyvern has a layout-sensitive textual concrete syntax (i.e.
+indentation is meaningful).
+This choice is not fundamental to our proposed mechanisms, but it will be useful for a class of examples that we will discuss later.
+Semantically, Wyvern consists of a \emph{module language} and a \emph{core language}.
+The module language is based directly on the Standard ML module language, with which we assume familiarity for the purposes of this proposal \cite{harper1997programming}.
+The core language (i.e.
+the language of types and expressions) is split into an \emph{external language} (EL) and an \emph{internal language} (IL).
+The semantics of EL constructs are specified by a type-directed translation to the much simpler IL.
+Notionally, this can be thought of as simply a shifting of the first stage of a type-directed compiler into the language specification \cite{tarditi+:til-OLD}.
+The internal language we will use is the polymorphic lambda calculus with binary product and sum types and recursive types (features like state and exceptions could also be included, but for simplicity, we will omit them) \cite{pfpl}.
+%Again, we will make these statements more precise as we continue.%The external language provides local type inference \cite{Pierce:2000:LTI:345099.345100}.
+% (for the purposes of this work).
+
+We will begin by describing Wyvern's mechanisms for expressing new forms of derived concrete syntax, assuming for the moment that the EL is otherwise specified in a standard way, in Section \ref{sec:syntax}.
+We then turn in Section \ref{sec:semantics} to a mechanism for introducing new semantic constructs -- in particular, new base types and operators -- into the EL.
+Constructs that are built in to languages like ML, like $n$-ary tuples/records, labeled sums and others, can in Wyvern be realized as modes of use of this mechanism.
+In both cases, the mechanisms can be broadly understood as novel forms of \emph{static metaprogramming}.
+Users define each new syntactic and semantic construct by writing code that generates terms and types.
+This code is  evaluated statically, in the course of typechecking EL terms.
+
+%Languages that are organized around mechanisms like these are colloquially known as \emph{extensible languages}.
+
+For each mechanism that we introduce, we will first demonstrate its expressive power by showing how constructs that are, or would need to be, built directly in to contemporary languages can in Wyvern be expressed in libraries.
+To demonstrate that these mechanisms are theoretically sound, we will then develop a formal specification and proofs of various important metatheoretic properties.
+The most interesting of these will be various \emph{modular reasoning principles} that guarantee that user-defined syntactic and semantic constructs can be reasoned about separately and then used together in any combination.
+These justify our classification of Wyvern as a \emph{modularly extensible programming language}.
+
+%Designing a modularly extensible language is a challenge because giving away too much control can easily make reasoning about program behavior difficult and lead to problems that arise only when extensions are used at the same time.
+For example, a library provider given free reign over syntax and semantics could introduce parsing conflicts, violate {type safety}, and  interfere with other extensions.
+We aim to design mechanisms that eliminate these possibilities, while providing a degree of expressive power sufficient to express a variety of constructs that are, or would need to be, built directly in to today's languages.
+%ll constructs that can be defined using a mechanism we will introduce can be reasoned about separately and composed arbitrarily.% With these mechanisms, we anticipate dialect formation being far less common.
+%Indeed, all of these problems have plagued previous  designs.
+
+% Common dialects of ML that suffice for \cite{harper1997programming}. %We will .
+
+   %Our challenge will be to design mechanisms that achieve these goals while also being powerful enough to express features that today require dialects.%Our challenge will be to design such extension mechanisms, while still retaining significant expressive power -- we wish to permit library-based expression of a range of features that today are, or would need to be, built into a system by its designers.%It can also make it more difficult to define new sorts of tools for working with the language because they cannot rely on there being a fixed abstract syntax (the canonical \emph{expression problem}).
 For programmers,
@@
 
-In this thesis, we aim to design mechanisms that handle these issues by embracing the already widely accepted notion that \emph{type constructors} (or more concisely, \emph{tycons}) can serve as the fundamental organizational unit of a programming system.
-We show how to delegate control over certain aspects of the concrete syntax, external type system and editor services to user-defined functions associated with type constructors, together forming what we call \emph{active type constructors}.
-This permits library-based realizations of features that today would need to be built into a system by its central designers.
-By layering these mechanisms over a fixed typed internal language and constraining these functions appropriately, we retain key  metatheoretic properties and arrive at powerful modular reasoning principles.
+%So our goal will be to design extension mechanisms that handle these metatheoretic and composition-related issues while still  permitting library-based expression of a range of features that today are, or would need to be, built into a system by its designers.
+%More specifically, we will show how to delegate control, in a methodical manner, to \emph{static functions}, i.e.
+user-defined functions written in a \emph{static language}.
+By layering these mechanisms over a fixed typed internal language, constraining the static language appropriately and validating the code it is used to generate, we will retain key  metatheoretic properties and arrive at powerful modular reasoning principles.
+%By associating these static functions with type constructors, forming \emph{active type constructors},  we will further show how they can enjoy a usage profile essentially identical to that of built in features.
 % while maintaining key metatheoretic properties and modular reasoning principles.
@@
 %To realize a new abstraction or system behavior, such experts can consider either a \emph{language-internal approach}, where they work within an existing language and distribute their solutions as libraries, or a \emph{language-external approach}, where they create a new, distinct programming system (often centered around what has come to be called a new \emph{domain-specific language} \cite{dsl}) or extend an existing system by some mechanism that is not part of the language itself, such as an extension mechanism supported by a {particular} compiler, editor or other tool.
-\subsection{Motivating Example: Regular Expressions}\label{regex}
-To make the use cases we seek to address more concrete, let us begin with a simple example that we will return to throughout this proposal.
-\emph{Regular expressions} are commonly used to capture patterns in strings (e.g.
-motifs in DNA sequences) \cite{Thompson:1968:PTR:363347.363387}.
-Programmers who work with regular expressions would benefit from features like these:
 
+%The result of this work will be the design of a \emph{modularly extensible programming language} called Wyvern.
+and end with a brief foray into editor services in Sec.
+\ref{sec:editor-services}.\todo{fix sections}
+
+\section{Modular Derived Syntax}\label{sec:syntax}
+Most contemporary computer programming languages specify a textual concrete syntax.
+Because it serves as the language's human-facing user interface, it is common practice to include  derived syntactic forms (colloquially, \emph{syntactic sugar}) that capture common idioms more concisely or naturally.
+% (i.e.
+considering cognitive dimensions \cite{green1996usability}).
+For example,  list syntax is built in to most dialects of ML, so that instead of having to write \lstinline{Cons(1, Cons(2, Cons(3, Nil)))}, a programmer can equivalently write \lstinline{[1, 2, 3]}.
+Many languages go further, building in syntax associated with other types of values, like vectors (SML), arrays (Ocaml), commands (Haskell) and syntax trees (Scala).
+%The desugaring from the latter to the former is specified by the language itself.
+%Typically, the language designer controls what forms of derived syntax are built in to the language.
+
+Rather than privileging  particular data structures, Wyvern instead exposes mechanisms that allow library providers to introduce new derived syntactic constructs on their own.
+%Lists need no special consideration from the language specification.
+%The purpose of this section is to motivate and then introduce Wyvern's syntax extension mechanisms.
+%For forms with a clear analogy to a form in Standard ML, we will assume the  semantics are analagous without providing details.
+To motivate our desire for this level of syntactic control, we begin in Sec.
+\ref{sec:examples} with the example of regular expression patterns expressed using abstract types, showing how the usual workaround of using strings to introduce patterns is not ideal.
+We then survey existing syntax extension mechanisms in Sec.
+\ref{sec:syntax-existing}, finding that they involve an unacceptable loss of modularity and other undesirable trade-offs.
+In Sec.
+\ref{sec:syntax-contributions}, we outline our proposed mechanisms and discuss how they resolve these issues.
+We conclude in Sec.
+\ref{sec:syntax-timeline} with a timeline for remaining work.
+
+\subsection{Motivating Example: Regular Expression Syntax}\label{sec:examples}
+Let us begin by taking the perspective of a \emph{regular expression} library provider.
+Recall that regular expressions are a common way to capture patterns in strings \cite{Thompson:1968:PTR:363347.363387}.
+We will assume that the abstract syntax of {patterns}, $p$, over strings, $s$, is specified as below:\[p ::= \textbf{empty} ~|~ \textbf{str}(s) ~|~ \textbf{seq}(p; p) ~|~ \textbf{or}(p; p) ~|~ \textbf{star}(p) ~|~ \textbf{group}(p)\]
+The most direct way to express this abstract syntax is by defining a recursive sum type \cite{pfpl}. Wyvern supports these as case types, which are analagous to datatypes in ML:
+
+\begin{lstlisting}[numbers=none]
+casetype Pattern
+  Empty
+  Str of string
+  Seq of Pattern * Pattern
+  Or of Pattern * Pattern
+  Star of Pattern
+  Group of Pattern
+\end{lstlisting}
+
+However, there are some reasons not to expose this representation of patterns directly to clients.
+First, patterns are usually identified up to congruence.
+For example, $\textbf{seq}(\textbf{empty}, p)$ is congruent to $p$.
+It would be useful if congruent patterns were indistinguishable from the perspective of client code.
+Second, it can be useful for performance reasons to maintain additional data alongside patterns (e.g.
+a corresponding finite automata here) without exposing this ``implementation detail'' to clients.
+Indeed, there are many ways to implement regular expression patterns, each with different performance trade-offs.
+For these reasons, a better approach is to define the following \emph{module type} (a.k.a.
+\emph{signature} in SML), where the type of patterns, \lstinline{t}, is held abstract.
+The client of any module \lstinline{P : PATTERN} can then identify patterns as terms of type \verb|P.t|.
+Notice that it exposes an interface otherwise isomorphic to the one available using a case type:
+
+\begin{lstlisting}[deletekeywords={case},numbers=none]
+module type PATTERN
+  type t
+  val Empty : t
+  val Str : string -> t
+  val Seq : t * t -> t
+  val Or : t * t -> t
+  val Star : t -> t
+  val Group : t -> t
+  val case : (
+    'a -> 
+    (string -> 'a) ->
+    (t * t -> 'a) ->
+    (t * t -> 'a) ->
+    (t -> 'a) ->
+    (t -> 'a) -> 
+    'a)
+\end{lstlisting}
+
+\paragraph{Concrete Syntax} The abstract syntax of patterns is too verbose to be used directly in all but the most trivial examples, so patterns are conventionally written using a more concise concrete syntax.
+For example, the concrete syntax \lstinline{A|T|G|C} corresponds to abstract syntax with the following encoding:
+\begin{lstlisting}[numbers=none,mathescape=|]
+P.Or(P.Str "SSTRAESTR", P.Or(P.Str "SSTRTESTR", P.Or(P.Str "SSTRGESTR", P.Str "SSTRCESTR")))
+\end{lstlisting}
+
+% \begin{figure}
+% \begin{center}
+% \includegraphics[scale=.55]{regex-palette.pdf}\\
+% $\Downarrow$ \text{(pressing \textbf{Enter})}\\
+% \includegraphics[scale=.55]{regex-code-generated.pdf}
+% \end{center}
+% \vspace{-20px}
+% \caption{An example of type-specific editor services, here shown for Java.}
+% \label{fig:regex-palette}
+% %\vspace{-10px}
+% \end{figure}
+
+%In a conventional \emph{monolithic} programming system, support for each of these features would need to be built into the language and tools.
+
+To encode the concrete syntax of patterns, regular expression library providers usually provide a utility function that converts strings to patterns.
+Because there may be many pattern implementations, the usual approach is to define a parameterized module (a.k.a.
+\emph{functor} in SML) defining such utility functions generically, like this:
+
+\begin{lstlisting}[numbers=none]
+module PatternUtil(P : PATTERN)
+  fun parse(s : string) : P.t
+    (* ... pattern parser here ... *)
+\end{lstlisting}
+This allows the client of any module \lstinline{P : PATTERN} to use the following definitions:
+\begin{lstlisting}[numbers=none]
+module PU = PatternUtil(P)
+let pattern = PU.parse
+\end{lstlisting}
+to construct patterns like this:
+\begin{lstlisting}[numbers=none]
+pattern "SSTRA|T|G|CESTR"
+\end{lstlisting}
+ %Again, none of our goals are comprehensively achieved.
+This approach is imperfect for several reasons:
 \begin{enumerate} 
-\item \textbf{Concrete syntax for pattern literals}.
-An ideal syntax would permit programmers to express patterns in the concise, conventional manner.
-For example, consider a bioinformatics  application: the BisI restriction enzyme cuts DNA when it sees the motif \texttt{GC\textit{N}GC}, where \texttt{\textit{N}} is any base.
-We would want to express such patterns with syntax like this, using curly braces to splice one pattern into another and parentheses to delimit captured groups:
+\item Our first problem is syntactic (or, one might say, aesthetic): string escape sequences conflict syntactically with pattern escape sequences. For example, the following will not be well-formed:
+\begin{lstlisting}[numbers=none,mathescape=|]
+let ssn = pattern "SSTR\d\d\d-\d\d-\d\d\d\dESTR"
+\end{lstlisting}
+In fact, when compiling an analagous term using SML of New Jersey (SML/NJ), we encounter the rather puzzling error message \verb|Error: unclosed string|.
+Using \verb|javac|, we encounter \verb|error: illegal escape character|.
+In a small lab study, we observed that this error was common, and nearly always initially misinterpreted by even experienced programmers who hadn't used regular expressions recently \cite{Omar:2012:ACC:2337223.2337324}.
+The workaround is to  double backslashes:
 \begin{lstlisting}[numbers=none]
-let N : Pattern = <SURLA|T|G|CEURL>
-let BisI : Pattern = <SURLGC({EURLNSURL})GCEURL>\end{lstlisting}
-%The cognitive load of reading and writing patterns is low and patterns are parsed once at compile-time. 
-Malformed patterns would result in intelligible {parser} errors.
-\item A \textbf{type system} that ensures that key invariants related to regular expressions are statically maintained. It might statically ensure that: 
-	\begin{enumerate}
-	\item only patterns, or properly escaped strings, are spliced into a pattern literal, to avoid splicing errors and injection attacks \cite{owasp2013, Bravenboer:2007:PIA:1289971.1289975}
-	\item out-of-bounds backreferences to a captured group do not occur \cite{spishak2012type}
-	\item string processing operations do not lead to a string that is malformed, when well-formedness can be captured as membership in a regular language \cite{fulton-thesis,HosoyaVouillonPierce2000ICFP}
+let ssn = pattern "SSTR\\d\\d\\d-\\d\\d-\\d\\d\\d\\dESTR"
+\end{lstlisting}
+
+\item Our second problem has an impact on both correctness and performance: pattern parsing does not occur until the pattern is evaluated.
+For example, the following malformed pattern will only trigger an error when this term is evaluated during the full moon: %Achieving this goal is an explicit goal of this proposal, so we are obviously not happy with this.
+
+\begin{lstlisting}[numbers=none]
+case moon_phase
+  Full => pattern "SSTR(GCESTR" (* malformedness not statically detected *)
+  _ => (* ... *)
+\end{lstlisting}
+Such issues can be found via testing, but empirical data gathered from open source projects suggests that there are many malformed regular expression patterns that are not detected by a project's test suite ``in the wild'' \cite{spishak2012type}.
+
+Parsing patterns at run-time also incurs a performance penalty.
+To avoid incurring it every time the pattern is encountered during evaluation, an appropriately tuned caching strategy must be introduced, increasing client complexity.% Regular expressions are often used across large datasets in scientific applications, so the absolute peformance penalty can be non-trivial.
+
+\item The final problem is that using strings to introduce patterns makes it more likely that programmers will use string concatenation to construct patterns derived from other patterns or from user input.
+For example, consider the following function:
+\begin{lstlisting}[numbers=none,escapechar=|]
+fun example_rx(name : string)
+  pattern (name ^ "SSTR: \\d\\d\\d-\\d\\d-\\d\\d\\d\\dESTR")
+\end{lstlisting}
+The (unstated) intent here is to treat \lstinline{name} as a pattern matching only itself, but this is not the observed behavior when \lstinline{name} contains special characters that are meaningful in patterns.
+The correct code is more verbose, again resembling abstract syntax:
+\begin{lstlisting}[numbers=none]
+fun example_fixed(name : string)
+  P.Seq(P.Str(name), P.Seq(pattern "SSTR: ESTR", ssn)) (* ssn as above *)
+\end{lstlisting}
+
+The mistake was the result of a programmer using a flawed heuristic, so it could be avoided with sufficient developer discipline. 
+The problem is that it is difficult to enforce this discipline mechanically.
+Both functions above have the same type and behave identically at many inputs, particularly those that would be expected during typical executions of the program (i.e.
+alphabetic names).
+In applications that query sensitive data, failures of this variety are observed to be both common and catastrophic from the perspective of security \cite{owasp2013}.
+Ideally, our library would be able to make it more difficult to inadvertently introduce subtle security bugs like this.
 \end{enumerate}
-%When an error is found, an intelligible error message is provided.
-%\item An \textbf{implementation} that partially or fully compiles known regular expressions into the efficient internal representation that will be used by the regular expression matching engine (e.g., a finite automata \cite{Thompson:1968:PTR:363347.363387}) ahead of time.
+% \begin{enumerate}
+% \item \textbf{Conventional Concrete Syntax.} Patterns must be written in exactly the conventional manner, shown in blue below.
+Curly braces serve as an ``unquote'' form to splice one pattern into another:
+% \begin{lstlisting}[numbers=none]
+% let N : Pattern = /SURLA|T|G|CEURL/
+% let BisI : Pattern = /SURLGC{EURLNSURL}GCEURL/
+% let twoDigits : Pattern = /SURL\d\dEURL/\end{lstlisting}
+
+% %The choice of delimiter, i.e. matching forward slashes here, is not important.
+% %The cognitive load of reading and writing patterns is low and patterns are parsed once at compile-time. 
+% \item \textbf{Compile-Time Parsing.} Pattern parsing must occur at compile-time, i.e.
+it must not incur run-time cost and malformed patterns, e.g.
+\lstinline{/SURLGC(EURL/}, must result in {compile-time} errors no matter where they appear in the program.
+% \item \textbf{Security By Default.} Strings must not be treated as patterns, nor used idiomatically to construct patterns.
+For example, the function below must be ill-typed because \lstinline{name} is not of type \lstinline{Pattern}:
+% \begin{lstlisting}[numbers=none]
+% fun example(name : string) : Pattern
+%   let twoDigits : Pattern = /SURL\d\dEURL/
+%   /SURL{EURLnameSURL}: {EURLtwoDigitsSURL}EURL/
+% \end{lstlisting}
+%     It is also acceptable for \lstinline{example("SSTR[a-z]ESTR")} to be equivalent to \lstinline{/SURL\[a-z\]: \d\d/}, i.e. strings are treated as patterns matching only themselves. 
+% It is \emph{not} acceptable for \lstinline{example("SSTR[a-z]ESTR")} to be equivalent to \lstinline{/SURL[a-z]: \d\d/}.
+
+%     This is a security issue because if \lstinline{name} is derived from user input, such confusion could lead to \emph{injection attacks}, which are both common and catastrophic when using regular expression libraries that fail to achieve these goals, discussed below \cite{owasp2013}.
+%   %\item A more advanced semantics might ensure that out-of-bounds backreferences to a captured group do not occur \cite{spishak2012type}.
+For example, we would want \verb|BisI|, above, to have the more precise type \lstinline{Pattern(1)}, i.e.
+a pattern with one captured group.
+% % \end{enumerate}
+% %When an error is found, an intelligible error message is provided.
+% %\item An \textbf{implementation} that partially or fully compiles known regular expressions into the efficient internal representation that will be used by the regular expression matching engine (e.g., a finite automata \cite{Thompson:1968:PTR:363347.363387}) ahead of time.
 In most languages, this compilation step occurs at run-time, even if the pattern is fully known at compile-time, thereby introducing performance overhead into programs.
@@
 %By performing this step ahead-of-time, these dangers can be avoided.
-\item \textbf{Editor services} to support syntax highlighting, consulting relevant documentation, interactive testing and pattern extraction from example strings.
-\end{enumerate}
+% %\item A range of \textbf{editor services} that support syntax highlighting for patterns, retrieval of relevant documentation, interactive testing, and pattern extraction from example strings have been shown to be helpful for programmers when working with complex regular expressions \cite{ACC_VLHCC}.
+An example of some of these is shown in Figure \ref{fig:regex-palette}: the editor service shown helps programmers generate correct code, here via the Java standard library's implementation of regular expressions (discussed below).
+% \end{enumerate}
+% Let us consider the ways that providers of regular expression libraries commonly choose to define the type \verb|Pattern|: as synonymous to \lstinline{string}, as a recursive sum type, or as an abstract type.
 
-%In a conventional \emph{monolithic} programming system, support for each of these features would need to be built into the language and tools. 
-No system today provides built-in support for all of the features enumerated above in their strongest form, so library providers must leverage  general-purpose mechanisms.
-%Unfortunately, it is impossible to completely define the syntax and the semantics referenced above using contemporary mechanisms, so library providers need to compromise.
+% \vspace{-5px}
+% \paragraph{Patterns as Strings} The most na\"ive choice is to define \lstinline{Pattern} as synonymous to \lstinline{string}:
 
-The most common strategy is to ask clients to introduce patterns as strings, deferring their  parsing, compilation and use to run-time.
-This provides only a partial approximation to feature 1 due to clashes between string escape sequences and regular expressions (parsing errors due to these clashes have been observed to be difficult for even experienced programmers to diagnose \cite{Omar:2012:ACC:2337223.2337324}).
-Moreover, none of the static guarantees can be provided, leading to run-time exceptions (common even in well-tested code \cite{spishak2012type}), and mistakes that are not caught even at run-time, some of which can lead to security vulnerabilities due to injection attacks \cite{owasp2013}.
-It also introduces performance overhead due to run-time parsing and compilation of patterns, and redundant run-time checks.
-Staging these computations can shift some of the errors and cost to compile-time, albeit with some additional semantic and syntactic complexity, but the issues related to the fact that a regular expression is not a string (e.g.
-syntactic clashes and injection attacks) persist.% parsing can move some errors to run-time, albeit at additional syntactic cost, but it also does not straightforwardly handle splicing-related issues, as we will discuss later.
+% \begin{lstlisting}[numbers=none]
+% type Pattern = string
+% \end{lstlisting}
+% The only reason to consider this definition is that it allows us to approximate pattern syntax using standard string literal syntax:
 
-A more semantically justifiable approach is to dispense with string representations entirely and directly work with an encoding of regular expression patterns using  general-purpose constructs like sums and  products (e.g.
-as exposed by functional datatypes or objects), or using an abstract type.
-This provides feature 2a while sacrificing feature 1 (such an encoding affords only an approximation of the verbose abstract syntax of patterns, not their concrete syntax).
-Advanced invariants like those described in features 2b and 2c are not directly tracked by such encodings and require solutions beyond those commonly found in modern simply-typed languages (we will return to this  in Sec.
-\ref{att}).%Explicitly constructing a regular expression by applying datatype constructors is unacceptably verbose.
+% \begin{lstlisting}[numbers=none]
+% let N : Pattern = "SSTRA|T|G|CESTR"
+% \end{lstlisting}
+% To splice one pattern into another, we can perform string concatenation:
+% \begin{lstlisting}[numbers=none,escapechar=|]
+% let BisI : Pattern = "SSTRGCESTR" ^ N ^ "SSTRGCESTR"
+% \end{lstlisting}
+% or use string interpolation (as in Scala or SML/NJ using quotation/antiquotation \cite{SML/Quote}):
+% \begin{lstlisting}[numbers=none,mathescape=|]
+% let BisI : Pattern = s"SSTRGC${ESTRNSSTR}GCESTR"
+% \end{lstlisting}
 
-None of these approaches address the important issue of tool support (feature 3).
-A number of tools are available online that provide services ranging from simple syntax highlighting to interactive testing and explanations of complex patterns \cite{regexr} and pattern extraction from examples (e.g.
-\cite{_txt2re:_????}).
-We have found that programmers benefit substantially from their use \cite{Omar:2012:ACC:2337223.2337324}.
-Unfortunately, evidence suggests that tools that must be accessed externally are difficult to discover and are thus used infrequently \cite{Murphy-Hill:2011:PIE:1958824.1958888,Campbell:2008:DRT:1636642.1636651,Omar:2012:ACC:2337223.2337324}.
-We have found that they are also less usable than editor-integrated tools %(e.g.
-\cite{IntelliJRegexp})
-because they require the programmer to switch contexts and cannot make use of the code context \cite{Omar:2012:ACC:2337223.2337324}.
-Working with high-level abstractions like these in a ``low-level'' editor is thus awkward and error-prone, just like attempting to work with high-level abstractions directly in terms of an encoding.
-For these reasons,   we see editor services as within the scope of an abstraction specification.
-\subsection{Language-External Approaches}\label{external-approaches}
-\begin{figure}
-\begin{center}
-\includegraphics[scale=.48]{approaches.pdf}
-\end{center}
-\vspace{-20px}
-\caption{\small (a) When taking a language-external approach, new features are packaged together into separate languages and tools.
-(b) When taking a language-integrated approach, there is one extensible host language and the compile-time and edit-time logic governing new constructs is expressed within ``active libraries''.}
-\label{approaches}
-%\vspace{-10px}
-\end{figure}
+% Though this comes close, goal 1 is not truly achieved because the syntax of strings can conflict with the syntax for patterns.
 
-In situations, like these, where a library-based approach is not satisfying, providers must take a \emph{language-external approach}, either by developing a new system dialect (supported by \emph{compiler generators} \cite{brooker1963compiler}, \emph{language workbenches} \cite{erdweg2013state}, \emph{DSL frameworks} \cite{fowler2010domain}, or simply by forking an existing codebase), or by using an extension mechanism for a {particular} compiler\footnote{Compilers that modify, or allow modification of, the semantics of their base language, rather than simply permitting semantics-preserving optimizations, should be considered a pernicious means for creating new language dialects.
-That is, some programs that purport to be written in C, Haskell or Standard ML are actually written in compiler-specific dialects of these languages.}, editor or other tool.
-For example, a researcher interested in providing the regular expression related features just described (let us refer to these collectively as \texttt{R}) might design a new system with built-in support for them, perhaps basing it on an existing system containing some general-purpose features (\texttt{G}).
-A different researcher developing a new language-integrated parallel programming abstraction (\texttt{P}) might  take the same approach.
-A third researcher, developing a type system for reasoning about units of measure (\texttt{Q}) might again do the same.
-This results in a collection of distinct systems, as diagrammed in Figure \ref{approaches}a.
-Although often justified as ``proofs of concept'', this is a rather weak notion because no abstraction will be used entirely in isolation and such a construction gives us no rigorously justifiable  reason to believe that it can safely and naturally coexist with others specified or implemented in the same way.
-That is, one must either use the system containing features \texttt{G+R}, \texttt{G+P} or \texttt{G+Q}.
-There is no system containing \texttt{G}, \texttt{R}, \texttt{P} and \texttt{Q} in other combinations, and merging the systems containing each separately can be non-trivial because there can be serious interference and safety issues, as we will discuss at length throughout this thesis.
-% More specifically, there is no guarantee of \textbf{orthogonality} or even \textbf{interoperability}.% This has limited the broad adoption of these kinds of innovations.%This latter method couples the semantics of the feature to the implementation details of a particular tool.
+% Goal 2 is not achieved because parsing occurs only at run-time when we match a string against a pattern (worse, every time we do so unless the implementation uses an effective caching strategy).
+
+% Goal 3 is also not achieved because the following code typechecks:
+% \begin{lstlisting}[numbers=none,mathescape=|]
+% fun example_bad_1(name : string) : Pattern
+%   let twoDigits : Pattern = "SSTR\\d\\dESTR"
+%   s"SSTR${ESTRnameSSTR}-${ESTRtwoDigitsSSTR}"ESTR
+% \end{lstlisting}
+% but \lstinline{example_bad_1("SSTR[a-z]ESTR")} evaluates to \lstinline{"SSTR[a-z]-\\d\\dESTR"}.
+
+% \paragraph{Patterns as Recursive Sums} A more semantically justifiable strategy is to define \lstinline{Pattern} as a recursive sum type  \cite{pfpl}.
+Wyvern supports these in the form of case types, which are analagous to datatypes in ML:
+
+% \begin{lstlisting}[numbers=none]
+% casetype Pattern
+%   Empty
+%   Str of string
+%   Seq of Pattern * Pattern
+%   Or of Pattern * Pattern
+%   Star of Pattern  
+% \end{lstlisting}
+
+% More generally, we can hold the representation of patterns abstract while exposing an isomorphic interface to clients by defining an abstract type in a {module} opaquely ascribed the following {module type} (a.k.a.
+\emph{signature} in SML): % exposes the constructors and case analysis as functions:
+
+% \begin{lstlisting}[deletekeywords={case},numbers=none]
+% module type PATTERN
+%   type t
+%   val Empty : t
+%   val Str : string -> t
+%   val Seq : t * t -> t
+%   val Or : t * t -> t
+%   val Star : t -> t
+%   val case : (
+%     'a -> 
+%     (string -> 'a) ->
+%     (t * t -> 'a) ->
+%     (t * t -> 'a) ->
+%     (t -> 'a) ->
+%     'a)
+% \end{lstlisting}
+% The client of any module \lstinline{P : PATTERN} can define \verb|Pattern| as synonymous to \verb|P.t|: 
+% \begin{lstlisting}[numbers=none]
+% type Pattern = P.t
+% \end{lstlisting}
+
+%The first one might not be found incorrect without rigorous testing with unexpected inputs.
+Goal 3 expresses the ideal of designing a language where security does not require considering such subtle issues extralinguistically.%This is precisely why injection attacks are so difficult to detect.
+
+% The dialect of Standard ML implemented by SML/NJ, recognizing that directly using constructors is often less than ideal, provides a facility for quotation/antiquotation \cite{SML/Quote}, which if introduced into Wyvern could be used as follows:
+% \begin{lstlisting}[numbers=none,escapechar=|,mathescape=|]
+% fun example(p : string)
+%   let twoDigits = P.qparse `SQT\d\dEQT`
+%   P.qparse `SQT^(EQTP.Str(p)SQT)-^(EQTtwoDigitsSQT)EQT`
+% \end{lstlisting}
+
+% Here, backticks are used to indicate a quoted term within which carets serve as an antiquote operator.
+This still represents a  compromise against goal 1.
+The most serious problem is that, once again, parsing does not occur until run-time.
+There are also  conflicts between quotation syntax and pattern syntax, i.e.
+\verb|^| conventionally means character set negation inside patterns, which conflicts with antiquotation here.
+This mechanism also does not support antiquotation at more than one type, so we could not define the antiquote operator to insert the \verb|Str| constructor automatically, as suggested in goal 3.
+%For syntax that necessary involves antiquotation at multiple types (e.g.
+concrete syntax for a programming language with multiple syntactic sorts, HTML which might include Javascript and CSS, and others), this
+
+\subsection{Existing Approaches}\label{sec:syntax-existing} These problems with string-oriented approaches to concrete syntax arise in many similar scenarios \cite{TSLs}, motivating  research on more direct syntax extension mechanisms \cite{Bravenboer:2007:PIA:1289971.1289975}.
+
+The simplest such mechanisms are those where each new syntactic form is described by a single equation.
+For example, the surface language of Coq (called Gallina) includes such a mechanism \cite{Coq:manual}.
+A theoretical account of such mechanisms has been developed by Griffin \cite{5134}.
+Unfortunately, these mechanisms are not able to express pattern syntax in the conventional manner for various reasons.
+For example, sequences of characters should not be parsed as identifiers.
+
+Syntax extension mechanisms based on context-free grammars like  Sugar* \cite{erdweg2013framework}, Camlp4 \cite{ocaml-manual} and many others are more expressive, and would allow us to directly introduce pattern syntax into our base language's grammar.
+However, this is perilous because none of the mechanisms described thusfar guarantee \emph{syntactic composability}, i.e.
+as stated in the Coq manual, ``mixing different symbolic notations in [the] same text may cause serious parsing ambiguity''.
+If another library provider used similar syntax for a different variant of regular expressions, or for an entirely unrelated abstraction, then a client could not simultaneously use both libraries in the same scope.
+% Resolving such parsing amibiguities is left to each client of the library.
+
+In response to this problem, Schwerdfeger and Van Wyk have developed a modular analysis that accepts only syntax extensions that use a unique starting token and satisfy some subtle conditions on follow sets of base language non-terminals \cite{conf/pldi/SchwerdfegerW09}.
+However, simple starting tokens like \lstinline{pattern} cannot be guaranteed to be globally unique, so we would need to use a more verbose token like \lstinline{edu_cmu_wyvern_rx_pattern}.
+There is no simple, principled way to define scoped abbreviations for starting tokens because this mechanism is language-external.
+
+In any case, we must now decide how our newly introduced derived forms desugar to forms in our base language, which in this case requires determining which module the constructors the desugaring uses will come from.
+Clearly, simply assuming that a module named \lstinline{P} satisying \lstinline{PATTERN} is in scope is a brittle solution.
+Indeed, we should expect that the extension mechanism actively prevents such capture of specific variable names to ensure that variables (including module variables) can be freely renamed.
+However, we note that such \emph{hygiene mechanisms} are only well-understood when performing term-to-term rewriting (as in Lisp-style \emph{macro systems}\todo{citation}) or simple equational systems like those found in Coq.
+For more flexible mechanisms, the issue is a topic of ongoing research (none of the grammar-based mechanisms described above enforce hygiene).
+
+Putting aside the question of hygiene, we can address the problem by requiring that the client explicitly identify the module the desugaring should use:
+\begin{lstlisting}[numbers=none]
+let N = edu_cmu_wyvern_rx_pattern[P] /A|T|G|C/
+\end{lstlisting}
+Syntactically, this is the best we can do using existing approaches.
+
+These approaches further suffer from a paucity of direct reasoning principles, i.e.
+the program can only be reasoned about post-desugaring.
+Given an unfamiliar piece of syntax, there is no simple method for determining what type it will have, or even for identifying which extension determines its desugaring, causing difficulties for both humans (related to code comprehension) and tools.
+
+% \begin{enumerate}
+% \item Given an unfamiliar piece of syntax, there is no simple method for determining what type it has, or to identify which extension determines its desugaring, hindering code comprehension.
+% \item It requires that the syntax be expressed as a grammar that can be consumed by a particular parser generator.
+Not all syntax  is most naturally expressed in this way, but there is no way to use a ``hand-rolled'' parser.
+% \end{enumerate}
+
+\subsection{Contributions}\label{sec:syntax-contributions}
+We propose designing a  syntax extension mechanism that addresses all of the problems just described. It consists of two constituent concepts:
+\begin{itemize}
+\item The concept that the mechanism is oriented around is the \emph{typed syntax macro} (TSM).
+%A TSM is applied to a \emph{delimited form}, which can contain  arbitrary syntax in its \emph{body}, and zero or more module or type parameters.
+Based on the body of the delimited form, the TSM statically computes an elaboration to a {base form}.
+For example, consider the following concrete term:
+\begin{lstlisting}[numbers=none]
+pattern[P] /SURLA|T|G|CEURL/
+\end{lstlisting}
+
+The TSM \lstinline{pattern} is being applied to a module parameter, \lstinline{P}, and a \emph{delimited form}, \lstinline{/SURLA|T|G|CEURL/}. This term elaborates \emph{statically} to the following:
+
+% import cs.cmu.edu/~wyvern/regex as R
+% let syntax pattern = R.pattern
+% let module P = R.P
+% The first line imports our top-level regular expression module (identified uniquely by a URI) and binds a shorter name, \lstinline{R}, to it.
+The second line binds a local name to the  TSM named \lstinline{pattern} that \lstinline{R} exports.
+The third line gives a shorter name, \lstinline{P}, to the module \lstinline{R.P : R.PATTERN}, as defined previously.
+The fourth line then applies \lstinline{pattern} to this module parameter and a form delimited by forward slashes.
+This causes it to elaborate \emph{statically} to the following base form:
+\begin{lstlisting}[numbers=none]
+P.Or(P.Str "SSTRAESTR", P.Or(P.Str "SSTRTESTR", P.Or(P.Str "SSTRGESTR", P.Str "SSTRCESTR")))
+\end{lstlisting}
+
+The TSM is defined as shown below:
+\begin{lstlisting}[numbers=none]
+syntax pattern[P : PATTERN] for P.t
+  fn (ps : ParseStream) => (* pattern parser here *)
+\end{lstlisting}
+We declare a module parameter (calling it \lstinline{P} here is unimportant, i.e.
+the TSM can be used with \emph{any} module satisfying module type \lstinline{PATTERN}), and the type clause \lstinline{for P.t}, which guarantees that any use of this TSM will either be ill-typed or elaborate to a term of type \lstinline{P.t}.
+The elaboration is computed by the compile-time action of the parse function defined in the indented block above.
+This function must be of type \lstinline{ParseStream -> Exp}, where the type \lstinline{ParseStream} gives the function access to the \emph{body} of the delimited form (in blue above) and the type \lstinline{Exp}  encodes the abstract syntax of the base language.
+Both types are defined in the Wyvern prelude, which is a set of definitions available ambiently.
+
+When the type of the term is known, e.g. due to a type annotation on the let binding, the module parameter \lstinline{P} can be inferred:
+\begin{lstlisting}[numbers=none]
+let N : P.t = pattern /SURLA|T|G|CEURL/
+\end{lstlisting}
+
+TSMs can be abbreviated using a let-binding style mechanism:
+\begin{lstlisting}[numbers=none]
+let syntax pat = pattern[P]
+pat /SURLA|T|G|CEURL/
+\end{lstlisting}
+
+TSMs can parse portions of the body of the delimited form as a base language term, to support splicing syntax:
+\begin{lstlisting}[numbers=none]
+let N = pat /SURLA|T|G|CEURL/
+let BisI = pat /SURLGC({EURLNSURL})GCEURL/
+\end{lstlisting}
+A hygiene mechanism ensures that only those portions of the elaboration derived from such spliced terms can refer to variables in the surrounding scope.
+
+Strings are never involved, so the attendant security issues described previously are easily avoidable. That is, the following
+
+\item To support an even more concise usage profile, Wyvern also  supports \emph{type-specific languages} (TSLs), which allow library providers to associate a TSM directly with a declared type.
+For example, the module \lstinline{P} above can associate \lstinline{pattern} with \lstinline{P.t}.
+Local type inference then determines which TSM is applied implicitly to handle a form not prefixed by a TSM name.
+For example, the following is equivalent to the above:
+\begin{lstlisting}[numbers=none]
+let N : P.t = /SURLA|T|G|CEURL/
+\end{lstlisting}
+\end{itemize}
+\subsubsection{Typed Syntax Macros (TSMs)}\label{sec:tsms}
+For simplicitly, let us begin with a typed syntax macro providing pattern syntax using the case type encoding above:
+\begin{lstlisting}[numbers=none]
+syntax pattern => Pattern = e_parser
+\end{lstlisting}
+The term \lstinline{e_parser}, elided here, must have type \lstinline{ParseStream -> Exp}, where \lstinline{ParseStream} classifies a sequence of characters and \lstinline{Exp} classifies a reified Wyvern expression.
+
+\subsubsection{Type-Specific Languages (TSLs)}\label{sec:tsls}
+
+\subsection{Timeline}\label{sec:syntax-timeline}
+SAC, ECOOP.
+
+TODO: module-parameterized TSMs
+
+\section{Extensible Semantics}\label{sec:extensible-semantics}
+
+\subsubsection{Example 2: Regular Strings}\label{sec:rstr}
+The example above deals with how regular expression patterns are  encoded, introduced and reasoned about.
+Going further, programmers may also benefit from a semantics where they can statically constrain strings to be within a particular regular language \cite{sanitation-psp14}.
+For example, a programmer might want to ensure that the arguments to a function that creates a database connection given a username and password are alphanumeric strings, ideally including this specification directly in the type of the  function:
+\begin{lstlisting}[numbers=none]
+type alphanumeric = rstring /SURL[A-Za-z0-9]+EURL/
+val connect : (alphanumeric * alphanumeric) -> DBConnection
+\end{lstlisting}
+
+This example requires that the language include a {type constructor}, \verb|rstring|, indexed by a statically known regular expression, written using the syntax described above.
+Ideally, we would be able to use standard string literal syntax for such regular strings as well, e.g.
+\begin{lstlisting}[numbers=none]
+let connection = connect("admin", "password")
+\end{lstlisting}
+
+To be useful, the language would also need a static semantics for standard operations on regular strings.
+For example, concatenating two alphanumeric strings  should result in an  alphanumeric string.
+This should not introduce additional run-time cost as compared to the corresponding operations on standard strings.
+Coercions that can be determined statically to be valid in all cases due to a language inclusion relationship should have trivial cost, e.g.
+coercions from alphabetic to alphanumeric strings.
+
+Regular strings might go beyond simply \emph{refining} standard string types (i.e.
+specifying verification conditions atop an existing semantics \cite{Freeman91}).
+For example, they might define an operation like \emph{captured group projection} that has no analog over standard strings:
+\begin{lstlisting}[numbers=none]
+let example : rstring /SURL(\d\d\d)-(\d\d\d\d)EURL/ = "555-5555"
+let group0 (* : rstring /\d\d\d/ *) = example#0
+\end{lstlisting}
+It should be possible to give this operation a cost of $\mathcal{O}(1)$.
+%From a language implementation perspective, this requires that different regular string types  have different representations, also not possible when using a refinement system over strings.
+%This is not possible if regular strings are defined using a refinement system atop a language with only strings.
+%This example requires introducing new operators into the system that did not already exist, and using an underlying represention that differs for different type indices.
+
+\subsubsection{Example 3: Labeled Products with Functional Update Operators}\label{sec:lprod}
+The simplest way to specify the semantics of product types is to specify only nullary and binary products \cite{pfpl}.
+However, in practice, many more general but also more complex variations on product types are built in to various dialects of ML and other languages, e.g.
+$n$-ary tuples, labeled tuples,
+records (identified up to reordering), 
+records with width and depth coercions \cite{Cardelli:1984:SMI:1096.1098} and records with functional update operators (also called \emph{extensible records}) \cite{ocaml-manual}.
+{The Haskell wiki notes that ''extensible records  are not implemented in GHC.
+The problem is that the record design space is large, and seems to lack local optima.
+[...] As a result, nothing much happens.'' \cite{GHCFAQ}}%,
+ %mutable fields \cite{ocaml-manual}, 
+%field delegation \cite{atlang-gpce14} \todo{gpce submission} 
+%and 
+% ``methods'' (i.e. pure objects) \cite{TSLs}
+
+We would ideally like to avoid needing the language designer to decide \emph{a priori} on just one privileged point in this large design space.
+Instead, the language designer might define only nullary and binary products, and include an extension mechanism that makes it possible for these other variations on products to be defined as libraries and  used together without conflict.
+For example, we would like to define the semantics of labeled products, which have labeled rows like records but maintain a row ordering like tuples, in a library.
+An example of a labeled product type (constructed by the type constructor \lstinline{lprod}) classifying conference papers might be:
+\begin{lstlisting}[numbers=none]
+type Paper = lprod {
+  title : rstring /SURL.+EURL/,
+   conf : rstring /SURL[A-Z]+ \d\d\d\dEURL/
+}
+\end{lstlisting}
+
+The row ordering should make it possible to introduce values of this type either with or without explicit labels, e.g.
+\begin{lstlisting}[numbers=none]
+fun make_paper(t : rstring /SURL.+EURL/) : Paper = {title=t, conf="EXMPL 2015"}
+\end{lstlisting}
+should be equivalent to
+\begin{lstlisting}[numbers=none]
+fun make_paper(t : rstring /SURL.+EURL/) : Paper = (t, "EXMPL 2015")
+\end{lstlisting}
+(note that we aim to be able to unambiguously re-use standard record and tuple syntax.)
+
+We should then be able to project out rows by providing a positional index or a label:
+\begin{lstlisting}[numbers=none]
+let test_paper = make_paper "Test Paper"
+test_paper#0
+test_paper#conf
+\end{lstlisting}
+
+We might also want our labeled tuples to support functional update operators. For example, an operator that dropped a row might be used like this:
+\begin{lstlisting}[numbers=none]
+let title_only (* : lprod {title : rstring /.+/} *) = test_paper.drop[conf]
+\end{lstlisting}
+An operation that added a row, or updated an existing row, might be used like this:
+\begin{lstlisting}[numbers=none]
+fun with_author(p : Paper, a : rstring /SURL.+EURL/) = p.ext(author=a)
+\end{lstlisting}
+% \subsection{Language-External Approaches}\label{external-approaches}
+
+% In situations like these where a library-based approach is not satisfactory, providers must today take a \emph{language-external approach}, specifying a new system dialect, reasoning about its metatheory monolithically and implementing it using a {compiler generator} \cite{brooker1963compiler}, {language workbench} \cite{erdweg2013state}, {DSL framework} \cite{fowler2010domain}, an extension mechanism for a {particular} compiler\footnote{Compilers that modify, or allow modification of, the semantics of their base language, rather than simply permitting semantics-preserving optimizations, should be considered a pernicious means for creating dialects.
+% %That is, some programs that purport to be written in C, Haskell or Standard ML are actually written in compiler-specific dialects of these languages.
+% }, editor or other tool, or by forking an existing system implementation.
+
+% For example, a researcher interested in providing the regular expression related features just described (let us refer to these collectively as \texttt{R}) might design a new system with built-in support for them, perhaps basing it on an existing system containing standard general-purpose features, \texttt{G} (e.g.
+the features of Standard ML).
+A different researcher developing a new language-integrated parallel programming abstraction, \texttt{P}, might  take the same approach (e.g.
+Manticore is such a system based on SML \cite{conf/popl/FluetRRSX07}).
+A third researcher, developing a type system for reasoning about units of measure (\texttt{Q}) might again do the same \cite{conf/cefp/Kennedy09}.
+This results in a collection of  systems, as diagrammed in Figure \ref{approaches}a.
+
+% Although these systems are valid ``proofs of concept'', this is a rather weak notion because no abstraction will be used entirely in isolation.
+Such an approach gives us no rigorously justifiable  reason to believe that it can safely and naturally be merged with others specified or implemented in the same way.
+That is, one must either use the system defining features \texttt{G+R}, \texttt{G+P} or \texttt{G+Q}.
+There is no system defining \texttt{G}, \texttt{R}, \texttt{P} and \texttt{Q} in other combinations.
+% and there is no formal, general mechanism for merging these systems.
+% We argue that, due to these problems of composition, taking a language-external approach to realizing a new feature should be considered harmful and avoided whenever possible.
+
+% \begin{figure}
+% \begin{center}
+% \includegraphics[scale=.48]{approaches.pdf}
+% \end{center}
+% \vspace{-20px}
+% \caption{\small (a) When taking a language-external approach, new features are packaged together into separate languages and tools.
+(b) We propose an approach where there is one extensible host language and the compile-time and edit-time logic governing new constructs is expressed within safely composable libraries.}
+% \label{approaches}
+% %\vspace{-10px}
+% \end{figure}
+%  %
+% % More specifically, there is no guarantee of \textbf{orthogonality} or even \textbf{interoperability}.% This has limited the broad adoption of these kinds of innovations.%This latter method couples the semantics of the feature to the implementation details of a particular tool.
 Because the use of one implementation entails a different semantics for the feature than another, the extended tool acts, \emph{de facto}, as a distinct system for our purposes.
 
-%\paragraph{Orthogonality} Features implemented by language-external means like these cannot be adopted individually, but instead are only available coupled to a fixed collection of other features.
+% %\paragraph{Orthogonality} Features implemented by language-external means like these cannot be adopted individually, but instead are only available coupled to a fixed collection of other features.
 This makes adoption more costly when these incidental features are  not desirable or are insufficiently developed (``toy languages''), or when the features bundled with a different language or tool are simultaneously desirable.
 
-Recent evidence indicates that this is one of the major barriers preventing research on programming languages from being driven into practice.
-For example, developers prefer high-level language-integrated parallel programming abstractions that provide  strong semantic guarantees when all else is equal \cite{cave2010comparing}, but library-based approximations are far more widely adopted because  ``parallel programming languages'' privilege only a few chosen  abstractions at the language level.
-This is problematic because different abstractions are seen as more appropriate in different situations \cite{Tasharofi:2013rc}.
-Moreover,  parallel programming is rarely the only concern relevant to clients outside of a classroom or research setting.
-Support for regular expressions, for example, would be simultaneously desirable for processing large amounts of genomic data in parallel, but using these features together in the same compilation unit would be difficult or impossible if implemented using language-external means.
+% % Recent evidence indicates that this is one of the major barriers preventing research  from being driven into practice.
+For example, developers prefer high-level language-integrated parallel programming abstractions that provide  stronger semantic guarantees  \cite{cave2010comparing}, but library-based approximations are far more widely adopted because the ``parallel programming languages'' privilege only a few  specialized abstractions at the language level.
+In contrast, it is widely acknowledged that  different specialized abstractions are more appropriate in different situations \cite{Tasharofi:2013rc}.
+Moreover,  parallel programming is rarely the only relevant specialized concern.
+Support for regular expressions as above would be simultaneously desirable for processing large amounts of genomic data in parallel.% but using these features together in the same compilation unit would be difficult or impossible if implemented using language-external means.
 Indeed, switching to a ``parallel programming language'' would likely make it \emph{more} difficult to use regular expressions, as these are likely to be less well-developed in a specialized language than in an established general-purpose language.% This intuition was perhaps most succinctly expressed by a participant in a recent study by Basili et al.
@@
 
-%\item Developing a new language and its associated tools places a significant development burden on providers who may wish only to promote a few core innovations, although tools like compiler generators, language workbenches and easy-to-extend tools can decrease this burden.
-%\item
+% %\item Developing a new language and its associated tools places a significant development burden on providers who may wish only to promote a few core innovations, although tools like compiler generators, language workbenches and easy-to-extend tools can decrease this burden.
+% %\item
 
-%Clients seem to prioritize the ability to choose different features for different portions of an application. 
-%If calling between languages were safe and easy, then using a variety of specialized languages and associated tools might be less problematic. In fact, s
-%Recognizing the limitations of relying on monolithic collection of primitives, some researchers have advocated instead for a model where multiple languages used within a single application, calling it the \emph{language-oriented approach} to software development \cite{languageoriented}.
+% %Clients seem to prioritize the ability to choose different features for different portions of an application. 
+% %If calling between languages were safe and easy, then using a variety of specialized languages and associated tools might be less problematic. In fact, s
+% %Recognizing the limitations of relying on monolithic collection of primitives, some researchers have advocated instead for a model where multiple languages used within a single application, calling it the \emph{language-oriented approach} to software development \cite{languageoriented}.
 
-\paragraph{Interoperability} Even in cases where, for each component of a software system, a programming system considered entirely satisfactory by its developers is available (e.g.
+% % \paragraph{Interoperability} Even in cases where, for each component of a software system, a programming system considered entirely satisfactory by its developers is available (e.g.
 a team goes through the trouble of implementing \verb|G+R+P| by reading papers about \verb|G+R| and \verb|G+P| and disentangling orthogonality-related issues), there remains a problem at any interface between  components written using a different combination of features.
@@
 
-One strategy taken by proponents of a {language-oriented approach} \cite{journals/stp/Ward94} to partially address the interoperability problem is to  target an established intermediate language and use its constructs as a common language for communication between components written in different languages.
+% % One strategy taken by proponents of a {language-oriented approach} \cite{journals/stp/Ward94} to partially address the interoperability problem is to  target an established intermediate language and use its constructs as a common language for communication between components written in different languages.
 Scala \cite{200464/IC} and F\# \cite{pickering2007foundations} are examples of prominent general-purpose languages that have taken this approach, and most DSL frameworks also rely on this strategy.
@@
 
-%This approach only works well when new languages consist of constructs that can also be expressed safely and almost as naturally in the common language.
-%But many of the most innovative constructs found in modern languages (often, those that justify their creation) are difficult to define in terms of existing constructs in ways that guarantee all necessary invariants are statically maintained and that do not require large amounts boilerplate code and run-time overhead.
-As a simple example with significant contemporary implications, F\#'s type system does not admit \verb|null| as a value for any type both defined and used within F\# code, but maintaining this sensible internal invariant still requires dynamic  checks because the stricter typing rules of F\# do not apply when F\# data structures are constructed by other languages on the Common Language Infrastructure (CLI) like C\# or SML.NET.
+% % %This approach only works well when new languages consist of constructs that can also be expressed safely and almost as naturally in the common language.
+% % %But many of the most innovative constructs found in modern languages (often, those that justify their creation) are difficult to define in terms of existing constructs in ways that guarantee all necessary invariants are statically maintained and that do not require large amounts boilerplate code and run-time overhead.
+% % As a simple example with significant contemporary implications, F\#'s type system does not admit \verb|null| as a value for any type both defined and used within F\# code, but maintaining this sensible internal invariant still requires dynamic  checks because the stricter typing rules of F\# do not apply when F\# data structures are constructed by other languages on the Common Language Infrastructure (CLI) like C\# or SML.NET.
 This is not an issue exclusive to intermediate languages that make regrettable choices regarding \verb|null|, however.
@@
 Moreover, it also cannot naturally consume F\# data structures, despite having a rather similar syntax and semantics in most ways (both languages directly descend from ML).
-%In Scala, traits that have default method implementations are difficult to implement from Java or other JVM languages and the workaround can break if the trait is modified \cite{scalatraitinterop}. 
-%In some cases, desirable features must be omitted entirely due to concerns about interoperability.
+% %In Scala, traits that have default method implementations are difficult to implement from Java or other JVM languages and the workaround can break if the trait is modified \cite{scalatraitinterop}.
+% %In some cases, desirable features must be omitted entirely due to concerns about interoperability.
 F\#, for example, aimed to retain source compatibility with Ocaml code, but due to the need for bidirectional interoperability with CLI languages, it does not support features like polymorphic variants, modules or functors \cite{ocaml-manual} because they have no apparent analogs in the type system of the CLI.
-%\end{itemize}
+% %\end{itemize}
 
-\subsection{Language-Integrated Approaches}\label{language-integrated-approaches}
-We argue that, due to these problems with orthogonality and interoperability, taking a language-external approach to realizing a new feature should be considered harmful and avoided whenever possible.
-The goal of the research being proposed here is to design \emph{language-integrated extension mechanisms} that give providers the ability to define, within libraries, new features that, like those above, have previously required central planning,
-%\footnote{One might compare today's programming systems to  {centrally-planned} economies, whereas extensible\- systems more closely resemble modern market economies.
-Our safety constraints serve a role analagous to market regulation.
-We leave further development of this analogy to the reader.}
-so that language-external approaches are less frequently necessary, as illustrated in Figure \ref{approaches}b. 
-Such libraries have been called \emph{active libraries}  \cite{activelibraries} in that they are not passive clients of features already available in the system.
-Features implemented within active libraries can be imported individually, unlike features implemented by external means, giving us a potential means to avoid the problems of orthogonality and interoperability.
+% %\subsection{Language-Integrated Approaches}\label{language-integrated-approaches}
 
-We must proceed with caution, however: critical issues having to do with {safety} must be overcome before language-integrated extension mechanisms can be introduced into a system.
-If too much control over  these core features of the system is given  to developers, the system may become quite unreliable.
-%For example, an extension could weaken important metatheoretic guarantees previously provided by the system. 
-Type safety, for example, may not hold if the static and dynamic semantics of the language can be modified or extended arbitrarily from within libraries.
-Furthermore, even if extensions can be shown not to cause such problems in isolation, there may still be conflicts between extensions that could weaken their semantics, leading to subtle problems that only appear when two extensions are used together, thwarting any attempts to reason modularly about particular programs and the system as a whole.
+% %Such libraries have been called \emph{active libraries}  \cite{activelibraries}.
+%Features implemented within active libraries can be imported individually, unlike features implemented by external means, giving us a potential means to avoid the problems of orthogonality and interoperability just described.
+
+% % We must proceed with caution, however: critical issues having to do with {safety} must be overcome before language-integrated extension mechanisms can be introduced into a system.
+If too much control over  these core features of the system is given  to developers, the system's metatheory may become weak.
+% % %For example, an extension could weaken important metatheoretic guarantees previously provided by the system. 
+% % For example, parsing ambiguities might arise if the syntax can be modified arbitrarily.
+Type safety  may not hold if the static and dynamic semantics of the language can be modified or extended arbitrarily from within libraries.
+Furthermore, even if extensions can be shown not to cause such problems in isolation, there may still be conflicts between extensions that could weaken their semantics, leading to subtle problems that only appear when two extensions are used together.
 %As a simple example, if two active libraries introduce the same syntactic form but back it with differing (but individually valid) semantics, this ambiguity  would only manifest itself when both libraries were imported within the same scope.
 %Resolving these kinds of ambiguities requires significantly more expertise with parser technology than using the syntax itself does.
-These issues have plagued previous attempts to design language-integrated extensibility mechanisms.
-We will briefly review some of these attempts below, then return to our approach.% To prevent them, our mechanisms will organize  extension logic around types to guarantee that extensions are both safe in isolation and also safely composable in any combination.
+% % These issues have plagued previous attempts to design language-integrated extensibility mechanisms.\todo{I can include a more detailed related work section if requested.}% We will briefly review some of these attempts below, then return to our approach.% To prevent them, our mechanisms will organize  extension logic around types to guarantee that extensions are both safe in isolation and also safely composable in any combination.
 
- %This represents a minimalist approach to system design -- the conventional distinction between built-in and user-defined constructs is blurred and most features of the system are orthogonally implemented as {libraries}, rather than by the maintainers of the system.
+%  %This represents a minimalist approach to system design -- the conventional distinction between built-in and user-defined constructs is blurred and most features of the system are orthogonally implemented as {libraries}, rather than by the maintainers of the system.
 
-%The mechanisms we describe will do so primarily by delimiting the scope of an extension to expressions of a single user-defined type or family of types.
+% %The mechanisms we describe will do so primarily by delimiting the scope of an extension to expressions of a single user-defined type or family of types.
 
-%This can be thought of as a more pernicious form of the conflict that arises when two globally-accessible constructs are given the same name.
+% %This can be thought of as a more pernicious form of the conflict that arises when two globally-accessible constructs are given the same name.
 n languages without universal namespacing mechanisms (e.g.
@@
 
-%The extension mechanism\todo{elaborate on safety requirements + tension between expressiveness and safety, merge with next paragraph}.
+% %The extension mechanism\todo{elaborate on safety requirements + tension between expressiveness and safety, merge with next paragraph}.
 must be expressive enough to allow users to associate rich run-time, compile-time and edit-time behaviors with user constructs directly, while being sufficiently restrictive to maintain the global safety properties of the language and system as a whole, and to ensure that constructs cannot interfere with one another.

### 2015-03-03 thesis-proposal/problem.tex
proposal update

@@
 % !TEX root = omar-thesis-proposal.tex
-\vspace{-25pt}
+%\vspace{-25pt}
 \section{Introduction}\label{motivation}
@@
 %%One might exp ``generality''
-Well-designed general-purpose programming languages are remarkably stable: they rarely need to be revised or forked into dialects, because programmers can express the constructs that they need just as well in orthogonal libraries.
+Ideally, a general-purpose programming language should only rarely need to be  extended or forked to form a dialect.
+Instead, programmers should be able to  express the constructs that they need in  libraries, as modes of use of existing constructs, and reason about their behavior modularly.
 %Instead, .
@@
 %Not all useful abstractions can be  realized, when considered comprehensively, as libraries.% The problem of achieving both simplicity and generality simultaneously with a single system remains unsolved.
-Abstraction mechanisms informed by type theory, like  the ML module system \cite{MacQueen:1984:MSM:800055.802036}, have put this ideal within reach. 
-Yet despite their expressive power, new {dialects} of languages like ML continue to arise. Why is this?
+Languages like ML, which features a powerful module system atop a core type structure rooted in the foundations of logic, appear to have put this ideal well within reach today. 
+Yet despite their expressive power, new {dialects} of languages like ML, and tools for creating such dialects, do continue to arise. What is motivating this?
 
@@
 %Why would a well-informed programmer consider forming a new dialect? 
-Perhaps the most common justification is that the \emph{syntactic cost} of a desirable construct is not  preserved when it is expressed using general-purpose mechanisms.
-%Syntactic cost is often assessed qualitatively \cite{green1996usability} (though quantitative metrics can also be defined).
-For example, the semantics of regular expressions can be expressed using abstract data types (which in ML are a mode of use of the module system), but as we will detail in Sec.
-\ref{sec:syntax}, this presents some syntactic difficulties.
-For programmers in problem domains  where regular expressions are common, e.g.
-bioinformatics, it is tempting to design a ``domain-specific'' dialect that differs only in that it includes derived syntax (i.e.
-``syntactic sugar'') for regular expression patterns.
-There is no shortage of tools that aid in the construction of such syntactic dialects (e.g.
-Camlp4 \cite{ocaml-manual}, SugarJ \cite{erdweg2011sugarj}, SugarHaskell \cite{erdweg2012layout} and many others).%Put another way, such embeddings may not preserve the cognitive cost associated with a construct (qualitative criteria are generally used to establish this \cite{green1996usability}).
+Perhaps the most common motivation is a concern about the \emph{syntactic cost} of expressing a construct of interest using general-purpose mechanisms (often assessed qualitatively \cite{green1996usability}, though quantitative metrics can be defined).
+For example, we consider regular expression patterns expressed using abstract data types (in ML,  a mode of use of the module system) in Sec.
+\ref{sec:syntax}.
+There is no semantic issue with this approach, but syntactically, it leaves much to be desired.
+In response to situations like this, it is not uncommon for providers to consider a \emph{syntactic dialect} of the language, i.e.
+one that specifies new concrete syntax by context-free elaboration to the existing language.
+For example, Ur/Web is a dialect of Ur (itself descended from ML) that builds in concrete syntax for SQL queries, HTML documents and other datatypes common in web programming \cite{conf/popl/Chlipala15}.
+Indeed, nearly all languages include some derived syntax for constructs that are otherwise expressed in the standard library, e.g.
+list syntax in Standard ML.
+Tools like Camlp4 \cite{ocaml-manual} and Sugar* \cite{erdweg2011sugarj,erdweg2013framework} are designed to lower the engineering costs of constructing syntactic dialects.
 
-Somewhat more rarely, dialects are motivated by the need to extend a general-purpose language with new derived semantic constructs, i.e.
-constructs that cannot be expressed as purely syntactic sugar, but where the general-purpose language is a suitable translation target.
-As a minimal example, consider System \textbf{F}, i.e.
-the polymorphic lambda calculus (perhaps the simplest general-purpose language) \cite{pfpl}.
-If we wanted to express record types in \textbf{F}, we could not do so simply with derived syntax.
-We could, however, develop a dialect that included the type and term operators relevant to records, specifying their static semantics directly but their dynamic semantics by translation to the polymorphic lambda calculus (using a standard Church-style encoding).\todo{couple more sentences} %Constructs specified by a type-directed translation like this can be considered \emph{derived semantic constructs}.
+%Many other data structures can similarly benefit from a decrease in syntactic cost.
+For example, we  will consider {regular expression patterns} encoded using abstract data types in Sec.
+\ref{sec:syntax}.
+%The syntactic cost of  such encodings, considered in several ways, can be reduced by using a dialect that builds in {syntactic sugar}  for regular expression patterns, and there are many tools that make constructing such syntactic dialects simple, e.g.
+Camlp4 \cite{ocaml-manual} and Sugar* \cite{erdweg2011sugarj,erdweg2013framework} (though as we will see, they come with costs of their own).
+%Put another way, such embeddings may not preserve the cognitive cost associated with a construct (qualitative criteria are generally used to establish this \cite{green1996usability}).
+
+More advanced dialects build in constructs that cannot simply be defined by a context-free elaboration to an existing language.
+For example, dialects of ML variously provide ``record-like'' constructs that support functional update and extension operators, mutable rows, width and depth coercions (often implicit) \cite{Cardelli:1984:SMI:1096.1098}, methods (i.e.
+function members that receive a self-reference implicitly), prototypic dispatch and various other embellishments.
+Ocaml builds in a class-based object system as well as various conveniences, e.g.
+logic for typechecking operations that use format strings, like $\mathtt{sprintf}$.
+ReactiveML builds in constructs for functional reactive programming \cite{mandel2005reactiveml}.
+ML5 builds in constructs for distributed programming \cite{Murphy:2007:TDP:1793574.1793585}.
+Manticore builds in parallel arrays \cite{conf/popl/FluetRRSX07}.
+Alice ML builds in futures and promises \todo{citation}.
+MLj builds in essentially the entire Java programming language, to support typesafe interoperation with existing Java code \cite{Benton:1999:IWW:317636.317791}.
+Indeed, perusal of the proceedings of any major programming languages conference will typically yield several more examples.
+Tools like proof assistants and logical frameworks are designed to support specifying and reasoning about dialects like these, and tools like compiler generators and language workbenches simplify the task of implementing them.
+
+%We will also discuss labeled products, which unlike records maintain a row ordering (this makes it possible to introduce them without explicit labels, but eliminate them using labels).
 
@@
 
-The problem with designing new language dialects in these situations is that a program written in one dialect cannot safely and idiomatically interface with libraries written in a different dialect.
-%At best, one can hope that the compilers for the two languages target a common intermediate language,
-To do so would require combining the two dialects into a single language, but there is no systematic method to do so, much less one that guarantees that important metatheoretic properties established about each dialect in isolation will be conserved.
-As a simple example, two dialects might be known to have an unambiguous concrete syntax in isolation, but if the syntax is na\"ively combined, ambiguities could easily arise.
+The problem with this practice of using language dialects as vehicles for new syntactic and semantic constructs is that it is, in an important sense,  anti-modular: a program written in one dialect cannot, in general, safely and idiomatically interface with libraries written in a different dialect.
+%At best, one can hope that the compilers for the decidedlytwo languages target a common intermediate language,
+This would require constructing a language where the constructs of both dialects are identically expressible, and where all the metatheoretic properties relevant to reasoning about program behavior are conserved.
+There can be no guarantee that such a language exists in general.
+As a very simple example, consider two dialects, one building in derived syntax for JSON (a popular data interchange format), the other using a similar syntax for finite mappings of some other type.
+Each is known to have an unambiguous concrete syntax in isolation, but when their syntax is na\"ively  combined, ambiguities arise.
+Combining the semantics of two languages, particularly when they are specified in very different ways, is even more of an art than a science, and subject to often subtle failures.
 %the ideas housed in one dialect are often only available to programmers willing to do without ideas housed in other dialects.
@@
 %A complex software system written in multiple distinct language dialects is far too unwieldy for it to be viable.
-Without strong modular reasoning principles to rely upon, it is difficult to construct large software systems and ecosystems. 
-%This means that it is difficult to construct large software systems and ecosystems from components written in different dialects. 
-The only reasonable approach for software engineers who must program ``in the large'' is to eschew constructs housed in dialects and settle on a single common language, where they can  rely on the guarantees provided by its type and module system.
+
+%It it clear that dialects are better rhetorical devices than practical engineering artifacts. 
+Given this, there are two approaches that the language design community can consider to put the ideas in dialects like those above into the hands of software engineers.
+One, exemplified by languages like Scala and Ocaml, is to progressively borrow new ideas from dialects as they emerge, integrating them into backwards compatible revisions.
+In taking this approach, power as well as responsibility to maintain metatheoretic guarantees is concentrated around a small group of language designers.
+This results in a large core language, yet because of a ``long tail'' effect, constructs that are situationally useful (or where there is disagreement) can still end up marginalized.
+Moreover, constructs that are included but turn out to be  flawed remain entrenched (and monopolize finite ``syntactic resources'', as we will discuss below).
+%This approach is antithetical to the ideal of a truly \emph{general-purpose language} described at the beginning of this section.
+
+% The ability to reason separately about libraries and import them freely is critical when constructing large software systems and participating in modern software ecosystems.
+If using a library requires a developer to directly modify the programming language they are using, detecting and resolving conflicts between constructs manually, then reestablish critical metatheoretic properties about the language followed by important invariants throughout their own code, the library will be essentially useless to them.
+% %This means that it is difficult to construct large software systems and ecosystems from components written in different dialects. 
+% From this perspective, it is hardly surprising when software engineers avoid dialects and settle on a single common language, where they can rely on the guarantees provided by its module system to program ``in the large''.
 %The designers of these languages may slowly adopt the most generally useful new ideas.
-This has left many useful constructs for programming ``in the small'' obscure.% It is safe to say that in contemporary times, language dialects should be considered harmful.
+% Unfortunately, this has left many useful constructs for programming ``in the small''  languishing in ``toy dialects''.% It is safe to say that in contemporary times, language dialects should be considered harmful.
 % It is little wonder that many programmers eschew esoteric language dialects in favor of languages that stopped evolving decades ago.
 
+% It is nearly always the case, however, that the dynamic semantics of these dialects can be defined by type-directed translation to the existing language.
+For example, ReactiveML builds in support for functional reactive programming, but its dynamic semantics can be implemented by type-directed translation to Standard ML \cite{mandel2005reactiveml}.
+In some cases, the semantics of the dialect are directly specified in this way.
+For example, F\# \cite{syme2012expert} introduces functional datatypes and the other constructs of the ML core language, defining them in terms of constructs available in the Common Intermediate Language (CIL), a class-based object-oriented language.
+Ocaml includes a class-based object system that can be defined by type-directed translation to Caml, a language building in little more than recursive product and sum types.
+
+{\Verse} aims to explore the polar opposite approach: a core language that builds in only minimal syntax and type structure \emph{a priori}.
+Instead, the core language is organized around metaprogramming mechanisms that give library providers the ability to introduce new concrete syntax and type structure of many designs, while imposing enough structure to ensure that the language maintains strong safety and modularity-related guarantees.
+This is far from a trivial matter.
+%\footnote{This is why languages  like Verse are often called \emph{extensible languages}, though this is somewhat of a misnomer.
+The chief characteristic of an extensible language is that it \emph{doesn't} need to be extended in situations where other languages would need to be extended.
+We have found that the relativity inherent in this definition is a point of confusion, so we will avoid this terminology and simply consider {\Verse} to be a (particularly expressive) programming language of the usual sort.}
+For example, if we simply conceived of the core language's semantics as being specified by a ``bag of rules'' (like ``paper'' specifications of languages often appear) and let library providers introduce new rules, one could easily define rules that violated type safety, interfered with invariants being maintained by rules defined in other libraries, or introduced non-determinism.
+There is also the question of how to allocate finite syntactic resources.
+For example, as noted above, the design space around record types is large.
+If we allow libraries to explore this design space, which library gets to determine the meaning of a form like \lstinlinep{\{label1=e1, label2=e2\}}?
+
+%This suggests a need for language-integrated mechanisms that allow library providers to modularly express syntactic and semantic constructs of the sort that today require new dialects, allowing for perhaps some low-cost syntactic compromises.
+Before we discuss how we address these problems, it is useful to acknowledge that completely eliminating the need for dialects would indeed be asking for too much: certain design decisions are fundamentally incompatible with others or require coordination across a language design.
+For example, constructs that require that the type system support subtyping could not be introduced uncompromisingly into a language that has no such semantic notion.
+Our interest is only in a subset of all constructs that can be specified in a mutually orthogonal manner by a semantics of a certain prototypic ``shape''.
+We will make this more specific below.
+%For example, it should be possible to express different record systems within a language specified like ML (or the simply typed lambda calculus).
+The challenge  is to give formal shape to these intuitions.
+%Our broad aim in this thesis is to design language mechanisms that take significant steps toward achieving this goal.
+
 \section{Contributions}
-Our aim in this thesis is to make dialect formation less frequently necessary by designing language mechanisms that give library providers the ability to express derived syntactic and semantic constructs from within the language, in a manner that ensures that these constructs can be reasoned about separately and composed arbitrarily.
+\subsection{Verse}
+To provide a coherent vehicle for our contributions, we introduce a new programming language called Verse.
+Let us briefly review its organization.
+Syntactically, Verse has a layout-sensitive textual concrete syntax (i.e.
+newlines and indentation are meaningful).
+This choice is not fundamental to our proposed mechanisms, but it will be useful for cleanly expressing a class of examples that we will discuss later.
+Semantically, Verse is organized into a \emph{module language} and a \emph{core language}.
+The module language is based directly on the Standard ML module language, with which we assume familiarity for the purposes of this proposal \cite{harper1997programming}.
+We will use it in an example in Section \ref{sec:syntax}.
+The core language, i.e.
+the language of types and expressions, is split into an \emph{external language} (EL) and an \emph{internal language} (IL).
+The EL  is specified by a type-directed translation semantics targeting the much simpler IL.
+Notionally, this can be thought of as simply shifting the first stage of a type-directed compiler directly into the language specification \cite{tarditi+:til-OLD}.
+More specifically, we specify a {bidirectionally typed translation semantics}, i.e.
+one where a distinction is made between \emph{type analysis} (the type is an ``input'') and \emph{type synthesis} (the type is an ``output''), for reasons that we will return to later.
 
+The mechanisms we introduce are relatively insensitive to the particular choice of IL.
+Choosing the semantics of the IL is the primary decision that a language designer leveraging these mechanisms would need to make.
+The only strict requirements are that the IL is type safe and supports parametric type abstraction.
+For our purposes, we will keep things simple by using the strictly evaluated polymorphic lambda calculus with nullary and binary product and sum types and recursive types, which we assume the reader has familiarity with and for which all the necessary metatheorems are already well-established \cite{pfpl}.
+Features like state, exceptions, concurrency primitives, scalars, arrays and other constructs characteristic of a first-stage compiler IL would also be included in practice, but we omit them when their inclusion would not meaningfully affect our discussion.
+
+\subsection{Modularly Metaprogrammable Concrete Syntax}
+There are three main mechanisms that we propose in this thesis.
+The first two have to do with concrete syntax, and are closely related, while the final one has to do with the type structure of the core language.
+\begin{itemize}
+\item \textbf{Typed Syntax Macros}.
+TODO: description + example (and a variant of them that further lowers syntactic cost in certain circumstances, \emph{type-specific languages}), introduced in Sec.
+\ref{sec:syntax}.
+\item \textbf{Tycon Structures}, introduced in Sec. \ref{sec:extensible-semantics}. 
+\end{itemize}
+For each mechanism that we introduce, we will first demonstrate its expressive power by showing how constructs that are, or would need to be, built directly in to contemporary languages can in Verse be expressed in libraries.
+To demonstrate that these mechanisms are theoretically sound, we will then develop a formal specification and proofs of various important metatheoretic properties.
+The most interesting of these will be various \emph{modular reasoning principles} that guarantee that user-defined constructs can be reasoned about separately and then used together in any combination, without the possibility of syntactic or semantic conflict.
+
+ %Only constructs for which the dynamic semantics can ultimately be defined by translation to the IL chosen can be expressed using the mechanisms we introduce (in practice, this would be exactly the same constraint necessary to guarantee that a compiler could implement a dialect specification).
+%Again, we will make these statements more precise as we continue.%The external language provides local type inference \cite{Pierce:2000:LTI:345099.345100}.
+% (for the purposes of this work).
 %Such mechanisms are colloquially known as \emph{extension mechanisms} because the sorts of constructs they are used to express are those that would otherwise require extensions to the language itself.
@@
 We will make this intuition more precise later.
+
+% Rather than building derived syntax for certain privileged data structures, e.g.
+list syntax, as is nearly ubiquitious in dialects of ML, Verse allows users to express new derived syntax on their own, associating it with their own data structures.
+To guarantee that syntactic ambiguities cannot arise, Verse controls which delimiters can be used around derived syntax, but otherwise imposes no
 
@@
 
-To provide a coherent vehicle for these mechanisms, we will introduce a new general-purpose language called Wyvern.
-Let us briefly summarize its organization.
-Syntactically, Wyvern has a layout-sensitive textual concrete syntax (i.e.
-indentation is meaningful).
-This choice is not fundamental to our proposed mechanisms, but it will be useful for a class of examples that we will discuss later.
-Semantically, Wyvern consists of a \emph{module language} and a \emph{core language}.
-The module language is based directly on the Standard ML module language, with which we assume familiarity for the purposes of this proposal \cite{harper1997programming}.
-The core language (i.e.
-the language of types and expressions) is split into an \emph{external language} (EL) and an \emph{internal language} (IL).
-The semantics of EL constructs are specified by a type-directed translation to the much simpler IL.
-Notionally, this can be thought of as simply a shifting of the first stage of a type-directed compiler into the language specification \cite{tarditi+:til-OLD}.
-The internal language we will use is the polymorphic lambda calculus with binary product and sum types and recursive types (features like state and exceptions could also be included, but for simplicity, we will omit them) \cite{pfpl}.
-%Again, we will make these statements more precise as we continue.%The external language provides local type inference \cite{Pierce:2000:LTI:345099.345100}.
-% (for the purposes of this work).
-
-We will begin by describing Wyvern's mechanisms for expressing new forms of derived concrete syntax, assuming for the moment that the EL is otherwise specified in a standard way, in Section \ref{sec:syntax}.
-We then turn in Section \ref{sec:semantics} to a mechanism for introducing new semantic constructs -- in particular, new base types and operators -- into the EL.
-Constructs that are built in to languages like ML, like $n$-ary tuples/records, labeled sums and others, can in Wyvern be realized as modes of use of this mechanism.
-In both cases, the mechanisms can be broadly understood as novel forms of \emph{static metaprogramming}.
-Users define each new syntactic and semantic construct by writing code that generates terms and types.
-This code is  evaluated statically, in the course of typechecking EL terms.
-
+ %Constructs that are built in to languages like ML, like $n$-ary tuples/records, labeled sums and others, can in Verse be realized as modes of use of this mechanism. 
 %Languages that are organized around mechanisms like these are colloquially known as \emph{extensible languages}.
 
-For each mechanism that we introduce, we will first demonstrate its expressive power by showing how constructs that are, or would need to be, built directly in to contemporary languages can in Wyvern be expressed in libraries.
-To demonstrate that these mechanisms are theoretically sound, we will then develop a formal specification and proofs of various important metatheoretic properties.
-The most interesting of these will be various \emph{modular reasoning principles} that guarantee that user-defined syntactic and semantic constructs can be reasoned about separately and then used together in any combination.
-These justify our classification of Wyvern as a \emph{modularly extensible programming language}.
+\subsection{Thesis Statement}
+In summary, we propose a thesis defending the following statement:
+\begin{quote}
+A programming language specified by a type-directed translation semantics can give library providers the ability to introduce derived concrete syntax and new external type structure, housed in constructs that can be reasoned about separately and used together in any combination.
+\end{quote}
 
@@
 
-%The result of this work will be the design of a \emph{modularly extensible programming language} called Wyvern.
+%The result of this work will be the design of a \emph{modularly extensible programming language} called Verse.
 and end with a brief foray into editor services in Sec.
@@
 
-Rather than privileging  particular data structures, Wyvern instead exposes mechanisms that allow library providers to introduce new derived syntactic constructs on their own.
+Rather than privileging  particular data structures, Verse instead exposes mechanisms that allow library providers to introduce new derived syntactic constructs on their own.
 %Lists need no special consideration from the language specification.
-%The purpose of this section is to motivate and then introduce Wyvern's syntax extension mechanisms.
+%The purpose of this section is to motivate and then introduce Verse's syntax extension mechanisms.
 %For forms with a clear analogy to a form in Standard ML, we will assume the  semantics are analagous without providing details.
@@
 We will assume that the abstract syntax of {patterns}, $p$, over strings, $s$, is specified as below:\[p ::= \textbf{empty} ~|~ \textbf{str}(s) ~|~ \textbf{seq}(p; p) ~|~ \textbf{or}(p; p) ~|~ \textbf{star}(p) ~|~ \textbf{group}(p)\]
-The most direct way to express this abstract syntax is by defining a recursive sum type \cite{pfpl}. Wyvern supports these as case types, which are analagous to datatypes in ML:
+The most direct way to express this abstract syntax is by defining a recursive sum type \cite{pfpl}. Verse supports these as case types, which are analagous to datatypes in ML:
 
@@
 % \paragraph{Patterns as Recursive Sums} A more semantically justifiable strategy is to define \lstinline{Pattern} as a recursive sum type  \cite{pfpl}.
-Wyvern supports these in the form of case types, which are analagous to datatypes in ML:
+Verse supports these in the form of case types, which are analagous to datatypes in ML:
 
@@
 
-% The dialect of Standard ML implemented by SML/NJ, recognizing that directly using constructors is often less than ideal, provides a facility for quotation/antiquotation \cite{SML/Quote}, which if introduced into Wyvern could be used as follows:
+% The dialect of Standard ML implemented by SML/NJ, recognizing that directly using constructors is often less than ideal, provides a facility for quotation/antiquotation \cite{SML/Quote}, which if introduced into Verse could be used as follows:
 % \begin{lstlisting}[numbers=none,escapechar=|,mathescape=|]
@@
 This function must be of type \lstinline{ParseStream -> Exp}, where the type \lstinline{ParseStream} gives the function access to the \emph{body} of the delimited form (in blue above) and the type \lstinline{Exp}  encodes the abstract syntax of the base language.
-Both types are defined in the Wyvern prelude, which is a set of definitions available ambiently.
+Both types are defined in the Verse prelude, which is a set of definitions available ambiently.
 
@@
 
-\item To support an even more concise usage profile, Wyvern also  supports \emph{type-specific languages} (TSLs), which allow library providers to associate a TSM directly with a declared type.
+\item To support an even more concise usage profile, Verse also  supports \emph{type-specific languages} (TSLs), which allow library providers to associate a TSM directly with a declared type.
 For example, the module \lstinline{P} above can associate \lstinline{pattern} with \lstinline{P.t}.
@@
 \end{lstlisting}
-The term \lstinline{e_parser}, elided here, must have type \lstinline{ParseStream -> Exp}, where \lstinline{ParseStream} classifies a sequence of characters and \lstinline{Exp} classifies a reified Wyvern expression.
+The term \lstinline{e_parser}, elided here, must have type \lstinline{ParseStream -> Exp}, where \lstinline{ParseStream} classifies a sequence of characters and \lstinline{Exp} classifies a reified Verse expression.
 
@@
 \section{Extensible Semantics}\label{sec:extensible-semantics}
+As a minimal example, consider System \textbf{F}, a.k.a.
+the polymorphic lambda calculus (perhaps the simplest general-purpose language) \cite{girard1971extension,Reynolds94anintroduction,pfpl}.
+It specifies type variables and two type constructors, written abstractly $\mathtt{arr}$ and $\mathtt{all}$:
+
+\[
+\begin{array}{llllll}
+\textbf{Sort} & & & \textbf{Abstract Form} & \textbf{Typeset Form} & \textbf{Description}\\
+\mathsf{Type} & \tau & ::= & t & t & \text{type variable}\\
+& & & \mathtt{arr}(\tau; \tau) & \tau \rightarrow \tau & \text{function type}\\
+& & & \mathtt{all}(t.\tau) & \forall t.\tau & \text{type quantification}
+\end{array}
+\]
+and value variables and four term constructors:
+\[
+\begin{array}{llllll}
+\textbf{Sort} & & & \textbf{Abstract Form} & \textbf{Typeset Form} & \textbf{Description}\\
+\mathsf{Exp} & \iota & ::= & x & x & \text{value variable}\\
+& & & \mathtt{lam}[\tau](x.\iota) & \lambda x{:}\tau.\iota & \text{value abstraction}\\
+& & & \mathtt{ap}(\iota; \iota) & \iota(\iota) & \text{application}\\
+& & & \mathtt{Lam}(t.\iota) & \Lambda t.\iota & \text{type abstraction}\\
+& & & \mathtt{Ap}[\tau](\iota) & \iota\mathtt{[}\tau\mathtt{]} & \text{type application}
+\end{array}
+\]
+
+If we wanted to express record types in \textbf{F}, derived syntax would not be enough (the static semantics of \textbf{F} provide no way to express row labels, amongst other issues).
+We could, however, develop a dialect that built in the type and term operators relevant to records, specifying its static and dynamic semantics as a type-directed translation to the polymorphic lambda calculus (using the standard Church-style encoding for records).
+More formally, assuming that metavariables $e$, $\sigma$ and $\Upsilon$ range over terms, types and typing contexts of our dialect, and $\iota$, $\tau$ and $\Gamma$  range over the terms, types and typing contexts of \textbf{F}, the main judgements in the specification of our dialect would take the following form:
+\[\begin{array}{ll}
+\textbf{Judgement} & \textbf{Pronunciation}\\
+\Upsilon \vdash e : \sigma \leadsto \iota & \text{Under typing context $\Upsilon$, term $e$ has type $\sigma$ and translation $\iota$.}\\
+\Upsilon \vdash \sigma \leadsto \tau & \text{Under typing context $\Upsilon$, type $\sigma$ is valid and has translation $\tau$.}\\
+\vdash \Upsilon \leadsto \Gamma & \text{Typing context $\Upsilon$ is valid and has translation $\Gamma$.}
+\end{array}\]
+\todo{couple more sentences} %Constructs specified by a type-directed translation like this can be considered \emph{derived semantic constructs}.
 
@@
 % \subsection{Language-External Approaches}\label{external-approaches}
+
+An important point, however, is that records, as well as all of these ``record-like'' constructs, can  be expressed by a \emph{type-directed translation} targeting an intermediate language that builds in only functions, nullary and binary products, recursive types and reference cells  \cite{pfpl} (and indeed, many compilers take advantage of facts like this to simplify the number of cases that must be considered downstream).
+%it is inaccurate to state that they are merely syntactic sugar over such a language.
+%Object systems are perhaps even more diverse across languages\todo{citation}.
+%Other dialects are justified by a desire to provide more special-purpose semantic constructs. For example, 
+Substantially more sophisticated examples require at most only a very slightly more capable intermediate language (the IL of popular virtual machines, even those with flaws like ubiquitous \texttt{null} references, e.g.
+JVM or CIL bytecode, are also sufficient if used strictly as translation targets).
+%Once again, these languages could all be specified by type-directed translation to a common intermediate language with comparatively few constructs , but generally not by a purely syntactic desugaring.
 

### 2015-03-24 thesis-proposal/problem.tex
Reworked proposal

--- 2015-02-06 thesis-proposal/problem.tex
@@
 % !TEX root = omar-thesis-proposal.tex
-\vspace{-25pt}
-\section{Introduction}\label{motivation}
+\vspace{-20pt}
+\section{Motivation}\label{motivation}
 % \begin{quote}\textit{The recent development of programming languages suggests that the simul\-taneous achievement of simplicity 
@@
 %%One might exp ``generality''
-Well-designed general-purpose programming languages are remarkably stable: they rarely need to be revised or forked into dialects, because programmers can express the constructs that they need just as well in orthogonal libraries.
-%Instead, .
-.
-%Indeed, a preponderence of new dialects is clear evidence against claims of generality.
+%Given a well-designed general-purpose programming language, programmers should be able to express the constructs that they need in libraries, as modes of use of existing constructs.
+New language dialects should be needed exceptionally rarely.
+
+Well-designed general-purpose programming languages give library providers the power to  modularly express a wide variety of useful constructs using a comparatively small set of primitives.
+%New primitives, and thus new dialects of these languages, should arise very rarely.
+%New primitives (and new \emph{dialects} of the language).
+%Extending a general-pandurpose language with new primitives, i.e.
+forming an extended \emph{dialect} of the language, should, given a sufficiently expressive such language, be exceedingly rare.
+%Indeed, a diversity of dialects around a language can be taken as evidence against claims of  generality.%, this can be taken as evidence against claims of its generality.
 %Not all useful abstractions can be  realized, when considered comprehensively, as libraries.% The problem of achieving both simplicity and generality simultaneously with a single system remains unsolved.
-Abstraction mechanisms informed by type theory, like  the ML module system \cite{MacQueen:1984:MSM:800055.802036}, have put this ideal within reach. 
-Yet despite their expressive power, new {dialects} of languages like ML continue to arise. Why is this?
+%Many languages claim to achieve this ideal. 
+%For example, 
+%If we measure the generality of a language by how frequently  dialects of the language emerge, 
+%all major contemporary languages seem to leave room for improvement. 
+%General-purpose programming languages are a dime a. 
+%Among the chief goals in the design of a general-purpose programming languages must achieve \emph{stability}: they  very rarely needs to be extended or forked into dialects by the user community surrounding it.
+Instead, its users are able to modularly  express a wide variety of useful constructs in libraries, using the comparatively small set of primitives that the language builds in.
+There has been substantial progress on many of the foundational semantic issues that are relevant to the design of general-purpose languages over the past  decades, much of it driven by advances in type theory.
+However, when more pragmatic language design issues are brought into consideration, a broadly agreeable set of primitives remains elusive, as evidenced by the diverse array of dialects that surround around all major con\-temporary languages.
+For example, let us consider the many dialects of ML.
+Though they generally agree on the importance of a powerful module system atop a core language based conceptually on the   polymorphic lambda calculus with sums, products and recursive types, they all introduce various syntactic and semantic extensions and tweaks of their own novel design (we discuss a number of examples below).
+Tools for  constructing new ``domain-specific'' language dialects also continue to proliferate.
+% The situation is similar across all major language lineages.
+Why are well-informed programmers and researchers so often tempted to construct  dialects of a language like ML? 
+%Let us investigate why this is.
+%What kinds of constructs are not expressible in languages like SML?
 
@@
 %Library implementations are not yet satisfying.
-%Why would a well-informed programmer consider forming a new dialect? 
-Perhaps the most common justification is that the \emph{syntactic cost} of a desirable construct is not  preserved when it is expressed using general-purpose mechanisms.
-%Syntactic cost is often assessed qualitatively \cite{green1996usability} (though quantitative metrics can also be defined).
-For example, the semantics of regular expressions can be expressed using abstract data types (which in ML are a mode of use of the module system), but as we will detail in Sec.
-\ref{sec:syntax}, this presents some syntactic difficulties.
-For programmers in problem domains  where regular expressions are common, e.g.
-bioinformatics, it is tempting to design a ``domain-specific'' dialect that differs only in that it includes derived syntax (i.e.
-``syntactic sugar'') for regular expression patterns.
-There is no shortage of tools that aid in the construction of such syntactic dialects (e.g.
-Camlp4 \cite{ocaml-manual}, SugarJ \cite{erdweg2011sugarj}, SugarHaskell \cite{erdweg2012layout} and many others).%Put another way, such embeddings may not preserve the cognitive cost associated with a construct (qualitative criteria are generally used to establish this \cite{green1996usability}).
-
-Somewhat more rarely, dialects are motivated by the need to extend a general-purpose language with new derived semantic constructs, i.e.
-constructs that cannot be expressed as purely syntactic sugar, but where the general-purpose language is a suitable translation target.
-As a minimal example, consider System \textbf{F}, i.e.
-the polymorphic lambda calculus (perhaps the simplest general-purpose language) \cite{pfpl}.
-If we wanted to express record types in \textbf{F}, we could not do so simply with derived syntax.
-We could, however, develop a dialect that included the type and term operators relevant to records, specifying their static semantics directly but their dynamic semantics by translation to the polymorphic lambda calculus (using a standard Church-style encoding).\todo{couple more sentences} %Constructs specified by a type-directed translation like this can be considered \emph{derived semantic constructs}.
+
+Perhaps the most common reason is simply that, for human {programmers}, the \emph{syntactic cost} of expressing a construct of interest using general-purpose primitives is not always ideal.
+%Syntactic cost is often assessed qualitatively \cite{green1996usability}, though quantitative metrics can be defined.
+For example,  we will consider regular expression patterns expressed using abstract data types (in ML, a mode of use of the module system) in Section \ref{sec:syntax}.
+This gives us precisely the semantics that we need, but the syntactic cost  is not quite ideal.
+In situations like this, library providers are often tempted to construct a \emph{derived syntactic dialect}, i.e.
+a dialect specified by context-free elaboration to the existing language, that introduces derived syntax specific to their library.
+For example, Ur/Web is a derived syntactic dialect of Ur (itself descended from ML) that builds in derived syntax for SQL queries, HTML elements and other datatypes common in web programming \cite{conf/popl/Chlipala15}.
+Indeed, nearly all major languages primitively build in derived syntax for constructs that are otherwise expressed in the standard library, e.g.
+derived list syntax in Standard ML (SML) \cite{harper1997programming,mthm97-for-dart}.
+Tools like Camlp4 \cite{ocaml-manual} and Sugar* \cite{erdweg2011sugarj,erdweg2013framework}, which we will return to later, help lower the engineering costs of constructing derived syntactic dialects, contributing to their proliferation.% contributing to their proliferation.
+
+%Many other data structures can similarly benefit from a decrease in syntactic cost.
+For example, we  will consider {regular expression patterns} encoded using abstract data types in Sec.
+\ref{sec:syntax}.
+%The syntactic cost of  such encodings, considered in several ways, can be reduced by using a dialect that builds in {syntactic sugar}  for regular expression patterns, and there are many tools that make constructing such syntactic dialects simple, e.g.
+Camlp4 \cite{ocaml-manual} and Sugar* \cite{erdweg2011sugarj,erdweg2013framework} (though as we will see, they come with costs of their own).
+%Put another way, such embeddings may not preserve the cognitive cost associated with a construct (qualitative criteria are generally used to establish this \cite{green1996usability}).
+
+More advanced dialects introduce  new type structure, going beyond what is possible by a context-free elaboration to an existing language.
+For example, SML goes beyond the simple nullary and binary product types common in simpler calculi, instead exposing record types.
+Other dialects have explored ``record-like'' primitives that support functional update and extension operators, width and depth coercions (sometimes implicit)%\cite{Cardelli:1984:SMI:1096.1098}
+, mutable rows, methods, prototypic dispatch and various other embellishments.
+Ocaml builds in a more  sophisticated object system, polymorphic variants and  conveniences like logic for typechecking operations that use format strings, e.g.
+$\mathtt{sprintf}$ \cite{ocaml-manual}.
+ReactiveML builds in primitives for functional reactive programming \cite{mandel2005reactiveml}.
+ML5 builds in primitives for distributed programming \cite{Murphy:2007:TDP:1793574.1793585}.
+Manticore builds in parallel arrays \cite{conf/popl/FluetRRSX07}.
+Alice ML builds in futures and promises \cite{AliceLookingGlass}.
+MLj builds in essentially the entire Java programming language, to support typesafe interoperation with existing Java code \cite{Benton:1999:IWW:317636.317791}.
+Indeed, perusal of the proceedings of any major programming languages conference will typically yield several more examples.
+Tools like proof assistants and logical frameworks are designed to support specifying and reasoning about dialects like these, and tools like compiler generators and language workbenches lower the engineering cost of implementing them.
+
+%We will also discuss labeled products, which unlike records maintain a row ordering (this makes it possible to introduce them without explicit labels, but eliminate them using labels).
 
@@
 
-The problem with designing new language dialects in these situations is that a program written in one dialect cannot safely and idiomatically interface with libraries written in a different dialect.
-%At best, one can hope that the compilers for the two languages target a common intermediate language,
-To do so would require combining the two dialects into a single language, but there is no systematic method to do so, much less one that guarantees that important metatheoretic properties established about each dialect in isolation will be conserved.
-As a simple example, two dialects might be known to have an unambiguous concrete syntax in isolation, but if the syntax is na\"ively combined, ambiguities could easily arise.
+The problem with this practice of using language dialects as vehicles for new syntactic and semantic primitives is that it is, in an important sense,  anti-modular: a program written in one dialect cannot, in general, safely and idiomatically interface with libraries written in a different dialect.
+%At best, one can hope that the compilers for the decidedlytwo languages target a common intermediate language,
+This would fundamentally require combining the two dialects into a single language.
+Left unconstrained, this is an ill-posed problem, i.e.
+given two dialects that are specified by judgements of arbitrary form, it is not clear what it would even mean to combine them.
+Even when the dialects are specified using a formalism that does provide some way to na\"ively combine two languages, there is generally no guarantee that the composition will conserve important syntactic and semantic properties.
+%There is no well-defined mechanism for constructing such a ``combined language'' in general.
+For example, consider two derived syntactic dialects, one building in derived syntax for JSON (a popular data interchange format), the other using a similar syntax for finite mappings of some other type.
+Though each is known to have an unambiguous concrete syntax in isolation, when their syntax is na\"ively  combined (e.g.
+by Camlp4), ambiguities arise.
+Any putative ``combined language'' must formally be considered a  distinct system for which one must derive all metatheorems of interest anew, guided only informally by those derived for the dialects individually.
+% would generally attempt to combine their semantics, particularly when they are specified in very different ways.
 %the ideas housed in one dialect are often only available to programmers willing to do without ideas housed in other dialects.
@@
 %A complex software system written in multiple distinct language dialects is far too unwieldy for it to be viable.
-Without strong modular reasoning principles to rely upon, it is difficult to construct large software systems and ecosystems. 
-%This means that it is difficult to construct large software systems and ecosystems from components written in different dialects. 
-The only reasonable approach for software engineers who must program ``in the large'' is to eschew constructs housed in dialects and settle on a single common language, where they can  rely on the guarantees provided by its type and module system.
-%The designers of these languages may slowly adopt the most generally useful new ideas.
-This has left many useful constructs for programming ``in the small'' obscure.% It is safe to say that in contemporary times, language dialects should be considered harmful.
-% It is little wonder that many programmers eschew esoteric language dialects in favor of languages that stopped evolving decades ago.
-
-\section{Contributions}
-Our aim in this thesis is to make dialect formation less frequently necessary by designing language mechanisms that give library providers the ability to express derived syntactic and semantic constructs from within the language, in a manner that ensures that these constructs can be reasoned about separately and composed arbitrarily.
-
-%Such mechanisms are colloquially known as \emph{extension mechanisms} because the sorts of constructs they are used to express are those that would otherwise require extensions to the language itself.
-%Defining a new construct then corresponds to the usual notion of language extension.
-We will make this intuition more precise later.
-
-%in languages without such mechanisms, expressing such constructs would require extensions to the language itself (a better qualifier might thus be \emph{extension-mitigating}).
-%We will make these notions more precise as we continue.
-%Thus, these mechanisms precisely capture the notions of language extension and language composition.
-
-%The mechanisms we propose are designed to decrease the number of constructs that need to be built in \emph{a priori} by a language designer.
-However, there are still some basic syntactic and semantic decisions that must unavoidably be made.
-
-To provide a coherent vehicle for these mechanisms, we will introduce a new general-purpose language called Wyvern.
-Let us briefly summarize its organization.
-Syntactically, Wyvern has a layout-sensitive textual concrete syntax (i.e.
-indentation is meaningful).
-This choice is not fundamental to our proposed mechanisms, but it will be useful for a class of examples that we will discuss later.
-Semantically, Wyvern consists of a \emph{module language} and a \emph{core language}.
-The module language is based directly on the Standard ML module language, with which we assume familiarity for the purposes of this proposal \cite{harper1997programming}.
-The core language (i.e.
-the language of types and expressions) is split into an \emph{external language} (EL) and an \emph{internal language} (IL).
-The semantics of EL constructs are specified by a type-directed translation to the much simpler IL.
-Notionally, this can be thought of as simply a shifting of the first stage of a type-directed compiler into the language specification \cite{tarditi+:til-OLD}.
-The internal language we will use is the polymorphic lambda calculus with binary product and sum types and recursive types (features like state and exceptions could also be included, but for simplicity, we will omit them) \cite{pfpl}.
-%Again, we will make these statements more precise as we continue.%The external language provides local type inference \cite{Pierce:2000:LTI:345099.345100}.
-% (for the purposes of this work).
-
-We will begin by describing Wyvern's mechanisms for expressing new forms of derived concrete syntax, assuming for the moment that the EL is otherwise specified in a standard way, in Section \ref{sec:syntax}.
-We then turn in Section \ref{sec:semantics} to a mechanism for introducing new semantic constructs -- in particular, new base types and operators -- into the EL.
-Constructs that are built in to languages like ML, like $n$-ary tuples/records, labeled sums and others, can in Wyvern be realized as modes of use of this mechanism.
-In both cases, the mechanisms can be broadly understood as novel forms of \emph{static metaprogramming}.
-Users define each new syntactic and semantic construct by writing code that generates terms and types.
-This code is  evaluated statically, in the course of typechecking EL terms.
-
-%Languages that are organized around mechanisms like these are colloquially known as \emph{extensible languages}.
-
-For each mechanism that we introduce, we will first demonstrate its expressive power by showing how constructs that are, or would need to be, built directly in to contemporary languages can in Wyvern be expressed in libraries.
-To demonstrate that these mechanisms are theoretically sound, we will then develop a formal specification and proofs of various important metatheoretic properties.
-The most interesting of these will be various \emph{modular reasoning principles} that guarantee that user-defined syntactic and semantic constructs can be reasoned about separately and then used together in any combination.
-These justify our classification of Wyvern as a \emph{modularly extensible programming language}.
-
-%Designing a modularly extensible language is a challenge because giving away too much control can easily make reasoning about program behavior difficult and lead to problems that arise only when extensions are used at the same time.
-For example, a library provider given free reign over syntax and semantics could introduce parsing conflicts, violate {type safety}, and  interfere with other extensions.
-We aim to design mechanisms that eliminate these possibilities, while providing a degree of expressive power sufficient to express a variety of constructs that are, or would need to be, built directly in to today's languages.
-%ll constructs that can be defined using a mechanism we will introduce can be reasoned about separately and composed arbitrarily.% With these mechanisms, we anticipate dialect formation being far less common.
-%Indeed, all of these problems have plagued previous  designs.
-
-% Common dialects of ML that suffice for \cite{harper1997programming}. %We will .
-
-   %Our challenge will be to design mechanisms that achieve these goals while also being powerful enough to express features that today require dialects.%Our challenge will be to design such extension mechanisms, while still retaining significant expressive power -- we wish to permit library-based expression of a range of features that today are, or would need to be, built into a system by its designers.%It can also make it more difficult to define new sorts of tools for working with the language because they cannot rely on there being a fixed abstract syntax (the canonical \emph{expression problem}).
-For programmers,
-%It can make it difficult to understand and reason about the type of an unfamiliar term, a critical facet of program comprehension (the \emph{typing discipline problem}).
-
-%So our goal will be to design extension mechanisms that handle these metatheoretic and composition-related issues while still  permitting library-based expression of a range of features that today are, or would need to be, built into a system by its designers.
-%More specifically, we will show how to delegate control, in a methodical manner, to \emph{static functions}, i.e.
-user-defined functions written in a \emph{static language}.
-By layering these mechanisms over a fixed typed internal language, constraining the static language appropriately and validating the code it is used to generate, we will retain key  metatheoretic properties and arrive at powerful modular reasoning principles.
-%By associating these static functions with type constructors, forming \emph{active type constructors},  we will further show how they can enjoy a usage profile essentially identical to that of built in features.
-% while maintaining key metatheoretic properties and modular reasoning principles.
-% By keeping the abstract syntax fixed and layering these mechanisms atop a fixed typed internal language (essentially lifting the first stages of compilation into the language), we avoid most aspects of the expression problem.
-%By delegating control to user-defined functions provided alongside user-defined type constructors (forming \emph{active type constructors}), we can give library providers substantially more control over these features of the system.
-%We structure each mechanism that we introduce as a \emph{bidirectional type system} combined with an  \emph{elaboration semantics} targeting a fixed typed internal language. 
-%By constraining these functions and enforcing critical abstraction barriers between extensions, we will ensure key metatheoretic properties of the language and system components and guarantee that extensions can be reasoned about modularly.
-%We call user-defined types that introduce new features into the system in this way \emph{active types}.
-% also highly expressive.
-%But taking a \emph{language-internal approach} to implementing a feature is the most practical.
-If a feature can be realized by creatively using existing language constructs and distributed as a library, clients face fewer barriers to adoption because it is easy to integrate library-based features into existing projects gradually and granularly and they leverage well-understood and well-developed mechanisms.
-% But taking this approach is often \emph{not} possible today %We call designs \emph{monolithic programming systems}.
-
-%To realize a new abstraction or system behavior, such experts can consider either a \emph{language-internal approach}, where they work within an existing language and distribute their solutions as libraries, or a \emph{language-external approach}, where they create a new, distinct programming system (often centered around what has come to be called a new \emph{domain-specific language} \cite{dsl}) or extend an existing system by some mechanism that is not part of the language itself, such as an extension mechanism supported by a {particular} compiler, editor or other tool.
-
-%The result of this work will be the design of a \emph{modularly extensible programming language} called Wyvern.
-and end with a brief foray into editor services in Sec.
-\ref{sec:editor-services}.\todo{fix sections}
-
-\section{Modular Derived Syntax}\label{sec:syntax}
-Most contemporary computer programming languages specify a textual concrete syntax.
-Because it serves as the language's human-facing user interface, it is common practice to include  derived syntactic forms (colloquially, \emph{syntactic sugar}) that capture common idioms more concisely or naturally.
+%It it clear that dialects are better rhetorical devices than practical engineering artifacts. 
+Due to this paucity of modular reasoning principles, language dialects are not practical artifacts for software development ``in the large''.
+%Large software projects and software ecosystems must pick a single language that does provide powerful modular reasoning principles and, to benefit from them, stay inside it.
+
+Dialects do, however, have an indirect impact on the practice of programming because the language designers that control comparatively popular languages, like Ocaml and Scala, can occasionally be convinced to incorporate  ideas from dialects  into backwards compatible language revisions.
+%These decisions are increasingly influenced by community processes, e.g.
+the Scala Improvement Process.
+%This approach concentrates power as well as responsibility over maintaining metatheoretic guarantees in the hands of a small group of language designers, though increasingly influenced by various community processes (e.g.
+the Scala Improvement Process).
+%Dialects thus serve the role of rhetorical vehicles for new ideas, rather than direct artifacts. 
+These major languages have thus become more expressive over time.
+However, this has come at a cost: they have also ballooned in size.
+Moreover, because there appears to be a ``long tail'' of potentially useful primitives (for derived syntactic primitives, our group has gathered initial data speaking to this \cite{TSLs}),  primitives that are only situationally useful, or that trigger aesthetic disagreements between different factions of the community, are still often left languishing in ``toy'' dialects.
+A conservative approach to incorporating new primitives is only sensible because once a primitive is introduced, it becomes entrenched as-is and monopolizes ``syntactic resources'', as we will discuss below.
+Recalling the words of  Reynolds, which are clearly as relevant today as they were nearly half a century ago \cite{Reynolds70}:%Because there is no data about how useful a construct is in practice until it is included in a language like this, decisions about which constructs to include are often informed by little more than intuition.
+%This approach is antithetical to the ideal of a truly \emph{general-purpose language} described at the beginning of this section.
+
+\begin{quote}\textit{The recent development of programming languages suggests that the simul\-taneous achievement of simplicity 
+and generality in language design is a serious unsolved 
+problem.}%\begin{flushright}-- John Reynolds (1970)\end{flushright}
+\end{quote}
+
+\newpage
+ There are two possible paths forward for the subset of the language design community interested in keeping general-purpose languages small and free of \emph{ad hoc} primitives.
+One, exemplified (arguably) by Standard ML, is to generally eschew the introduction of new primitives and settle on a set of primitives that sit at a ``sweet spot'' in the overall language design space, accepting that in some circumstances, these primitives trade away expressive power or have high syntactic cost.
+% that this implies.
+The other path forward is to press on in the search for even more general language primitives, i.e.
+primitives that are so expressive that they allow us to degrade a broad class of existing primitives, including those found in dialects, into modularly composable library constructs, where they can be evaluated on their individual merits by programmers.
+% This should render the construction of new dialects increasingly unnecessary.
+Such primitives do sometimes arise.
+For example, Ocaml recently introduced  ``generalized algebraic data types'' (GADTs), based on research on guarded recursive datatype constructors \cite{XiCheChe03}.
+Using GADTs, Ocaml was able to move some machinery for typechecking operations that use format strings, like \texttt{printf}, out of the language and into the standard library (though some machinery remains built in).
+Our broad aim will be to introduce a new general-purpose programming language called Verse\footnote{We distinguish Verse from Wyvern, which is the language referred to in prior publications about some of the work being proposed here, because Wyvern is a group effort evolving independently in some important ways.} that takes even more dramatic steps down this second path.
+%Similarly, it recently introduced ``open datatypes'', which subsume its previous more specialized exception type, and captures many use cases for .
+
+%Viewed ``dually'', one might equivalently ask for a language that builds in a core that is as small as possible, but provides expressive power comparable to languages with much larger cores.
+This is our goal in the work being proposed.
+
+\section{Proposed Contributions}
+
+Verse has a module system taken directly from SML, but its core language is organized in a novel manner around a minimal typed lambda calculus, called the Verse \emph{internal language}, introduced in Section \ref{sec:verse}.
+In lieu of typical {external} (i.e.
+programmer-facing) derived syntax and  type structure, Verse introduces two novel primitive constructs: \textbf{typed syntax macros} (TSMs),  in Section \ref{sec:syntax}, and \textbf{metamodules},  in Section \ref{sec:metamodules}.
+Briefly summarized, TSMs subsume the need for derived syntax specific to library constructs (e.g.
+list syntax in SML) by giving library providers static control over the parsing and elaboration of delimited segments of textual concrete syntax.
+Metamodules give library providers static hooks directly into the semantics of the Verse EL, allowing  library providers to, for example, define the type structure of records (and various ``record-like'' constructs, as discussed above) by type-directed translation to the internal language (which builds in only nullary and binary products).
+Both TSMs and metamodules can be understood as novel \emph{metaprogramming} primitives, because they involve static code generation.
+We will  also introduce a simple variant of each of these primitives that leverages Verse's support for local type inference to further reduce syntactic cost in certain common situations.
+
+The key challenge in the design of these primitives will come in ensuring that they are metatheoretically well-behaved, given that they aim to give library providers decentralized control over aspects of the language that have typically been under the monolithic control of the language designer.
+If we are not careful, many of the problems mentioned above as inherent to combining distinct language dialects could simply manifest themselves in the semantics of these primitives.\footnote{This is why languages  like Verse are often called \emph{extensible languages}, though this is somewhat of a misnomer.
+The chief characteristic of an extensible language is that it \emph{doesn't} need to be extended in situations where other languages would need to be extended.
+We will avoid this somewhat confusing terminology.} Our main technical contributions will come in showing how to address these problems in a principled manner.
+In particular, syntactic ambiguities will be impossible by construction and  we will validate the code  statically generated by TSMs and metamodules to maintain a hygienic type discipline and, most uniquely, powerful modular reasoning principles.
+Library providers will be able to reason about the constructs that they have defined in isolation, and library clients will be able to use them safely in any combination, without the possibility of conflict.\footnote{We will assume that simple naming conflicts can be avoided extrinsically in some suitable manner, e.g.
+by using a URI-based naming scheme as in the Java ecosystem.}
+
+\subsection{Thesis Statement}
+In summary, we propose a thesis defending the following statement:
+\begin{quote}
+A programming language can  give library providers the power to meta\-pro\-gram\-matic\-ally express new derived syntax and external type structure atop a minimal typed internal language, using  primitives that maintain a hygienic type discipline and modular reasoning principles.
+These  primitives are  expressive enough to subsume the need for a variety of primitives that are, or would need to be, built in to comparable contemporary languages.
+\end{quote}
+
+\subsection{Disclaimers}
+Before we continue, it may be useful to explicitly acknowledge that completely eliminating the need for dialects would indeed be asking for too much: certain design decisions are fundamentally incompatible with others or require coordination across a language design.
+We aim only to decrease the need for dialects.% out a larger design space within a single language, Verse.%a subset of constructs that can be specified by a semantics of a certain ``shape'' specified by Verse (we will make this more specific later).
+%There is nothing ``universal'' about Verse.
+
+It may also be useful to explicitly acknowledge that library providers could leverage the primitives we introduce   to define constructs that are in rather poor taste.
+We  expect that in practice, Verse will come with a standard library defining a carefully curated collection of standard constructs, as well as guidelines for advanced users regarding when it would be sensible to use the primitives we introduce (following the example of languages that support operator overloading or typeclasses, which also have the potential for ``abuse'').
+%The vast majority of programmers should not use the primitives that we introduce directly.
+
+Finally, we are not interested here in languages that feature full-spectrum dependent types, which blur the phase separation between compile-time and run-time, though we conjecture that the primitives we introduce could be introduced into languages like Gallina (the ``external language'' of the Coq proof assistant) with some modifications.
+Verse should be compared to languages that maintain a phase separation, like ML and Scala.
+%Our interest is in making sure that the standard library is no more privileged than any other library.
+% Our interest is not in creating a universal language, only one that is as expressive as reasonably possible.
+%only in a subset of all constructs that can be specified in a mutually orthogonal manner by a semantics of a certain prototypic ``shape''.
+We will make this more specific later.
+
+\section{Verse}\label{sec:verse}
+Let us begin by briefly summarizing how Verse is organized and specified.
+Verse consists of a \emph{module language} and a \emph{core language}.
+The module language is based directly on SML's module language,  which we assume familiarity with for the purposes of this work \cite{harper1997programming,MacQueen:1984:MSM:800055.802036}.
+We will give an example of its use in Section \ref{sec:syntax}, but do not plan to specify it in detail.
+
+The core language is organized much like the first stage of a type-directed compiler (e.g.
+the TIL compiler for Standard ML \cite{tarditi+:til-OLD}), consisting of a user-facing \emph{typed external language} (EL) specified by type-directed translation to a minimal \emph{typed internal language} (IL).
+The main judgements in the specification of the EL take the following form (we omit mention of various contexts for now):
+\\[1ex]
+$\begin{array}{ll}
+\textbf{Judgement} & \textbf{Pronunciation}\\
+e \Rightarrow \sigma \leadsto \iota & \text{External expression $e$ synthesizes external type $\sigma$ and has translation $\iota$.}\\
+e \Leftarrow \sigma \leadsto \iota & \text{External expression $e$ analyzes against external type $\sigma$ and has translation $\iota$.}\\
+\sigma~\mathtt{type} \leadsto \tau & \text{Static expression $\sigma$ is an external type with translation $\tau$.}
+\end{array}
+$\\[1ex]
+Note that the typing judgements are \emph{bidirectional}, i.e.
+there is a distinction between \emph{type synthesis} (the type is an ``output'') and \emph{type analysis} (the type is an ``input'') \cite{Pierce:2000:LTI:345099.345100}.
+This is to  allow us to explicitly specify how Verse's \emph{local type inference} works, and in particular, how it interacts with the primitives that we will introduce.
+Like Scala, we intentionally do not aim to support global type inference, both because it is not compatible with the primitives we introduce as described and because we believe that in practice, the cognitive cost of the error messages that arise outweigh its marginal utility relative to local type inference.
+Note also that, for concision, we will write ``expression'' and ``type'' without qualification to refer to  external expressions and types, respectively.
+
+% The reason why we refer to the syntactic sort $\sigma$ as a ``static expression'' will become more clear later.
+
+%The main novelty of our work is that the Verse EL builds in very little syntax or type structure \emph{a priori}.
+%Instead, \textbf{metamodules} and \textbf{typed syntax macros} (TSMs) are expressive enough to render this unnecessary.
+Although metamodules are in some sense prior to TSMs, we will describe TSMs first, temporarily assuming that the EL does build in standard type structure, for pedagogical purposes.
+
+Our design is largely insensitive to the details of the IL.
+The only strict requirements are that the IL is type safe and supports parametric type abstraction.
+For our purposes, we will keep things simple by using the strictly evaluated polymorphic lambda calculus with nullary and binary product and sum types and recursive types, which we assume the reader has familiarity with and for which all the necessary metatheorems are well-established (we follow \cite{pfpl}).
+The main judgements in the static semantics of the IL (again omitting contexts for now) thus take the following form:
+\\[1ex]
+$
+\begin{array}{ll}
+\textbf{Judgement} & \textbf{Pronunciation}\\
+\iota : \tau & \text{Internal expression $\iota$ has internal type $\tau$.}\\
+\tau~\mathtt{type} & \text{Internal type $\tau$ is valid.}
+\end{array}
+$\\
+
+\noindent
+The dynamic semantics can be specified as a transition system with  judgements of the following form:
+\\[1ex]
+$
+\begin{array}{ll}
+\textbf{Judgement} & \textbf{Pronunciation}\\
+\iota \mapsto \iota' & \text{Internal expression $\iota$ transitions to $\iota'$.}\\
+\iota~\mathtt{val} & \text{Internal expression $\iota$ is a value.}
+\end{array}
+$
+\\[1ex]
+The iterated transition judgement $\iota \mapsto^{*} \iota'$ is the reflexive, transitive closure of the transition judgement, and the evaluation judgement $\iota \Downarrow \iota'$ is derivable iff $\iota \mapsto^{*} \iota'$ and $\iota'~\mathtt{val}$.
+
+Primitive features like state, exceptions, concurrency primitives, scalars, arrays and others characteristic of a first-stage compiler IL would also be included in practice, and in some cases this would affect the shape of the internal semantics, but we omit them when their inclusion would not meaningfully affect our discussion.
+%Features like record types and labeled sum types, on the other hand, are not built in.
+
+\section{Modularly Metaprogrammable Textual Syntax}\label{sec:syntax}
+Verse, like most major contemporary programming languages, specifies a textual concrete syntax.
+We have chosen to specify a layout-sensitive textual concrete syntax (i.e.
+newlines and indentation are not ignored).
+This design choice is not  fundamental to our proposed contributions, but it will be useful for cleanly expressing a class of examples that we will discuss later.
+We will specify key aspects of its concrete syntax with an \emph{Adams grammar} \cite{Adams:2013:PPI:2429069.2429129} (such a specification for Wyvern, which has a very similar syntax, can be found in \cite{TSLs}), but for the purposes of this proposal, we will simply introduce Verse's concrete syntax by example as we go on.
+
+A programming language's concrete syntax serves as its human-facing user interface, so it is common practice to build in  derived forms (colloquially, \emph{syntactic sugar}) that capture common idioms more concisely or naturally.
 % (i.e.
 considering cognitive dimensions \cite{green1996usability}).
-For example,  list syntax is built in to most dialects of ML, so that instead of having to write \lstinline{Cons(1, Cons(2, Cons(3, Nil)))}, a programmer can equivalently write \lstinline{[1, 2, 3]}.
-Many languages go further, building in syntax associated with other types of values, like vectors (SML), arrays (Ocaml), commands (Haskell) and syntax trees (Scala).
+For example, derived list syntax is built in to most dialects of ML, so that instead of having to write \lstinline{Cons(1, Cons(2, Cons(3, Nil)))}, a programmer can equivalently write \lstinline{[1, 2, 3]}.
+Many languages go further, building in syntax associated with various other types of data, like vectors (SML/NJ), arrays (Ocaml), commands (Haskell), syntax trees (Scala), XML data (Scala, Ur/Web) and SQL queries (F\#, Ur/Web).
 %The desugaring from the latter to the former is specified by the language itself.
@@
 
-Rather than privileging  particular data structures, Wyvern instead exposes mechanisms that allow library providers to introduce new derived syntactic constructs on their own.
+Rather than privileging  particular library constructs with primitive syntactic support, Verse exposes primitives that allow library providers to introduce new derived syntax on their own.
 %Lists need no special consideration from the language specification.
-%The purpose of this section is to motivate and then introduce Wyvern's syntax extension mechanisms.
+%The purpose of this section is to motivate and then introduce Verse's syntax extension mechanisms.
 %For forms with a clear analogy to a form in Standard ML, we will assume the  semantics are analagous without providing details.
 To motivate our desire for this level of syntactic control, we begin in Sec.
-\ref{sec:examples} with the example of regular expression patterns expressed using abstract types, showing how the usual workaround of using strings to introduce patterns is not ideal.
-We then survey existing syntax extension mechanisms in Sec.
+\ref{sec:examples} with a representative example: regular expression patterns expressed using abstract data types.
+We show how the usual workaround of using strings to introduce patterns is not ideal.
+We then survey existing alternatives in Sec.
 \ref{sec:syntax-existing}, finding that they involve an unacceptable loss of modularity and other undesirable trade-offs.
@@
 We will assume that the abstract syntax of {patterns}, $p$, over strings, $s$, is specified as below:\[p ::= \textbf{empty} ~|~ \textbf{str}(s) ~|~ \textbf{seq}(p; p) ~|~ \textbf{or}(p; p) ~|~ \textbf{star}(p) ~|~ \textbf{group}(p)\]
-The most direct way to express this abstract syntax is by defining a recursive sum type \cite{pfpl}. Wyvern supports these as case types, which are analagous to datatypes in ML:
+The most direct way to express this abstract syntax is by defining a recursive sum type \cite{pfpl}.
+Verse supports these as case types, which are analagous to datatypes in ML (we will see how the type structure  of case types can in fact be expressed in libraries later):
 
@@
 However, there are some reasons not to expose this representation of patterns directly to clients.
-First, patterns are usually identified up to congruence.
-For example, $\textbf{seq}(\textbf{empty}, p)$ is congruent to $p$.
-It would be useful if congruent patterns were indistinguishable from the perspective of client code.
+First, patterns are usually identified up to their reduction to a normal form.
+For example, $\textbf{seq}(\textbf{empty}, p)$ has normal form $p$.
+It might be useful for patterns with the same normal form to be  indistinguishable from the perspective of client code.
 Second, it can be useful for performance reasons to maintain additional data alongside patterns (e.g.
-a corresponding finite automata here) without exposing this ``implementation detail'' to clients.
-Indeed, there are many ways to implement regular expression patterns, each with different performance trade-offs.
-For these reasons, a better approach is to define the following \emph{module type} (a.k.a.
-\emph{signature} in SML), where the type of patterns, \lstinline{t}, is held abstract.
+a corresponding finite automata) without exposing this ``implementation detail'' to clients.
+Indeed, there are many ways to represent regular expression patterns, each with different performance trade-offs.
+For these reasons, a better approach is to use \emph{abstract data types}.
+In Verse, as in ML, this is a mode of use of the module system.
+In particular, we can define the following \emph{module signature}, where the type of patterns, \lstinline{t}, is held abstract.
 The client of any module \lstinline{P : PATTERN} can then identify patterns as terms of type \verb|P.t|.
-Notice that it exposes an interface otherwise isomorphic to the one available using a case type:
+%Notice that it exposes an interface otherwise  to the one available using a case type:
 
 \begin{lstlisting}[deletekeywords={case},numbers=none]
-module type PATTERN
+signature PATTERN
   type t
@@
 \end{lstlisting}
+By holding the representation type of patterns abstract, the burden of proving that the case analysis function above cannot be used to distinguish patterns with the same normal form is localized to each module implementing this signature.
+The details are standard and not particularly interesting given our purposes, so we omit them.
 
 \paragraph{Concrete Syntax} The abstract syntax of patterns is too verbose to be used directly in all but the most trivial examples, so patterns are conventionally written using a more concise concrete syntax.
-For example, the concrete syntax \lstinline{A|T|G|C} corresponds to abstract syntax with the following encoding:
+For example, the concrete syntax \lstinline{A|T|G|C} corresponds to abstract syntax with the following much more verbose expression:
 \begin{lstlisting}[numbers=none,mathescape=|]
@@
 
-To encode the concrete syntax of patterns, regular expression library providers usually provide a utility function that converts strings to patterns.
-Because there may be many pattern implementations, the usual approach is to define a parameterized module (a.k.a.
-\emph{functor} in SML) defining such utility functions generically, like this:
+To express the concrete syntax of patterns, regular expression library providers usually provide a utility function that transforms strings to patterns.
+As just mentioned, there  may be many implementations of the \lstinline{PATTERN} signature, so the standard approach is to define a \emph{parameterized module} (a.k.a.
+\emph{functor} in SML) defining utility functions generically:
 
@@
 \begin{enumerate} 
-\item Our first problem is syntactic (or, one might say, aesthetic): string escape sequences conflict syntactically with pattern escape sequences. For example, the following will not be well-formed:
+\item Our first problem is syntactic: string escape sequences conflict syntactically with pattern escape sequences. For example, the following will not be well-formed:
 \begin{lstlisting}[numbers=none,mathescape=|]
@@
 \end{lstlisting}
-In fact, when compiling an analagous term using SML of New Jersey (SML/NJ), we encounter the rather puzzling error message \verb|Error: unclosed string|.
-Using \verb|javac|, we encounter \verb|error: illegal escape character|.
-In a small lab study, we observed that this error was common, and nearly always initially misinterpreted by even experienced programmers who hadn't used regular expressions recently \cite{Omar:2012:ACC:2337223.2337324}.
-The workaround is to  double backslashes:
+When compiling an analagous expression using \verb|javac|, we encounter the error message \verb|error: illegal escape character|.\footnote{When compiling an analagous expression using SML of New Jersey (SML/NJ), we encounter the seemingly inexplicable error message \texttt{Error: unclosed string}.} In a small lab study, we observed that this error was common, and nearly always initially misinterpreted by even experienced programmers who hadn't used regular expressions recently \cite{Omar:2012:ACC:2337223.2337324}.
+The inelegant workaround is to  double backslashes:
 \begin{lstlisting}[numbers=none]
@@
 
-\item Our second problem has an impact on both correctness and performance: pattern parsing does not occur until the pattern is evaluated.
-For example, the following malformed pattern will only trigger an error when this term is evaluated during the full moon: %Achieving this goal is an explicit goal of this proposal, so we are obviously not happy with this.
+Some languages, anticipating such modes of use, build in special string forms that leave escape sequences uninterpretted. For example, Ocaml allows the following, which is perhaps more acceptable:
+\begin{lstlisting}[numbers=none]
+let ssn = pattern {rx|SSTR\d\d\d-\d\d-\d\d\d\dESTR|rx}
+\end{lstlisting}
+
+\item The next problem is that pattern parsing does not occur until the pattern is evaluated at run-time.
+For example, the following malformed pattern will only trigger an error when this expression is evaluated during the full moon: %Achieving this goal is an explicit goal of this proposal, so we are obviously not happy with this.
 
@@
 \end{lstlisting}
-Such issues can be found via testing, but empirical data gathered from open source projects suggests that there are many malformed regular expression patterns that are not detected by a project's test suite ``in the wild'' \cite{spishak2012type}.
+Such issues can sometimes be found via testing, but empirical data gathered from open source projects suggests that there remain many malformed regular expression patterns that are not detected by a project's test suite ``in the wild'' \cite{spishak2012type}.
+
+We might instead attempt to treat the well-formedness of patterns constructed from strings as a static verification condition.
+Statically proving that this condition holds throughout a program is wildly difficult to automate in general, because it involves reasoning about arbitrary dynamic behavior.
+In the example above, we must know that the variable \lstinline{pattern} is equivalent to the function \lstinline{PU.pattern}.
+Moreover, if the argument were not written literally, e.g.
+we had written \lstinline{strconcat "SSTR(GESTR" "SSTRCESTR"}, we would need to be able to establish that this is equivalent to the string above.
+It is simple to come up with arbitrarily more complex examples that thwart any putative decision procedure, particularly for patterns that are dynamically constructed based on input to the program (further discussed below).
 
 Parsing patterns at run-time also incurs a performance penalty.
-To avoid incurring it every time the pattern is encountered during evaluation, an appropriately tuned caching strategy must be introduced, increasing client complexity.% Regular expressions are often used across large datasets in scientific applications, so the absolute peformance penalty can be non-trivial.
-
-\item The final problem is that using strings to introduce patterns makes it more likely that programmers will use string concatenation to construct patterns derived from other patterns or from user input.
+To avoid incurring it every time the pattern is encountered (e.g.
+for each item in a very large dataset), one must be careful to stage the computation appropriately, or use an appropriately tuned caching strategy.
+%This too is difficult if a portion of the pattern is dynamically generated.
+% Regular expressions are often used across large datasets in scientific applications, so the absolute peformance penalty can be non-trivial.
+
+\item The final problem is that using strings to construct patterns encourages programmers to use string concatenation inappropriately to construct patterns derived from user input.
 For example, consider the following function:
 \begin{lstlisting}[numbers=none,escapechar=|]
-fun example_rx(name : string)
+fun example_bad(name : string)
   pattern (name ^ "SSTR: \\d\\d\\d-\\d\\d-\\d\\d\\d\\dESTR")
 \end{lstlisting}
-The (unstated) intent here is to treat \lstinline{name} as a pattern matching only itself, but this is not the observed behavior when \lstinline{name} contains special characters that are meaningful in patterns.
-The correct code is more verbose, again resembling abstract syntax:
+The (unstated) intent here is to treat \lstinline{name} as a sub-pattern matching only itself, but this is not the observed behavior when \lstinline{name} contains special characters that mean something else in patterns.
+The correct code again resembles abstract syntax:
 \begin{lstlisting}[numbers=none]
@@
 
-The mistake was the result of a programmer using a flawed heuristic, so it could be avoided with sufficient developer discipline. 
-The problem is that it is difficult to enforce this discipline mechanically.
+In applications that query sensitive data, failures of this variety can lead to \emph{injection attacks}.
+These are observed to be both common and catastrophic from the perspective of security \cite{owasp2013}.
+The attack vector above was, of course, the fault of the programmer using a flawed heuristic, so it could have been avoided with sufficient discipline.
+The problem is again that it is difficult to detect violations of this discipline automatically.
 Both functions above have the same type and behave identically at many inputs, particularly those that would be expected during typical executions of the program (i.e.
-alphabetic names).
-In applications that query sensitive data, failures of this variety are observed to be both common and catastrophic from the perspective of security \cite{owasp2013}.
-Ideally, our library would be able to make it more difficult to inadvertently introduce subtle security bugs like this.
+alphabetic names), so testing often misses the problem.
+Formally proving that such problems do not arise involves reasoning about complex run-time data flows.
+
+ %Ideally, our library would be able to make it more difficult to inadvertently introduce subtle security bugs like this.
 \end{enumerate}
@@
 
-\subsection{Existing Approaches}\label{sec:syntax-existing} These problems with string-oriented approaches to concrete syntax arise in many similar scenarios \cite{TSLs}, motivating  research on more direct syntax extension mechanisms \cite{Bravenboer:2007:PIA:1289971.1289975}.
-
-The simplest such mechanisms are those where each new syntactic form is described by a single equation.
-For example, the surface language of Coq (called Gallina) includes such a mechanism \cite{Coq:manual}.
-A theoretical account of such mechanisms has been developed by Griffin \cite{5134}.
-Unfortunately, these mechanisms are not able to express pattern syntax in the conventional manner for various reasons.
-For example, sequences of characters should not be parsed as identifiers.
-
-Syntax extension mechanisms based on context-free grammars like  Sugar* \cite{erdweg2013framework}, Camlp4 \cite{ocaml-manual} and many others are more expressive, and would allow us to directly introduce pattern syntax into our base language's grammar.
-However, this is perilous because none of the mechanisms described thusfar guarantee \emph{syntactic composability}, i.e.
-as stated in the Coq manual, ``mixing different symbolic notations in [the] same text may cause serious parsing ambiguity''.
+\subsection{Existing Approaches}\label{sec:syntax-existing} These difficulties with string-oriented approaches to concrete syntax are not particular to regular expression patterns, but rather arise in many similar scenarios, motivating much research on reducing the need for run-time string parsing \cite{Bravenboer:2007:PIA:1289971.1289975}.
+Existing approaches take one of two approaches: \emph{syntax extension} or \emph{static term rewriting}.
+
+\subsubsection{Syntax Extension}
+One common approach is to use a system that allows users to directly extend the syntax of a language with new derived syntax, in this case for patterns.
+
+The simplest such systems are those where each new syntactic form is described by a single rewrite rule.
+For example, the external language of Coq (called Gallina) includes such a system \cite{Coq:manual}.
+A formal account of such systems has been developed by Griffin \cite{5134}.
+Unfortunately, these systems are not able to express pattern syntax in the conventional manner.
+For example, sequences of characters  can only be parsed as identifiers using these systems, rather than as characters in a pattern.
+Syntax extension systems based on context-free grammars like  Sugar* \cite{erdweg2013framework}, Camlp4 \cite{ocaml-manual} and many others are more expressive, and would allow us to directly introduce pattern syntax into our core language's grammar.
+
+However, this is a perilous approach because none of the systems described thusfar guarantee \emph{syntactic composability}, i.e.
+as the Coq manual states: ``mixing different symbolic notations in [the] same text may cause serious parsing ambiguity''.
 If another library provider used similar syntax for a different variant of regular expressions, or for an entirely unrelated abstraction, then a client could not simultaneously use both libraries in the same scope.
+Properly considered, every combination of extensions introduced by one of these mechanisms creates a \emph{de facto} derived syntactic dialect of the language.
 % Resolving such parsing amibiguities is left to each client of the library.
 
-In response to this problem, Schwerdfeger and Van Wyk have developed a modular analysis that accepts only syntax extensions that use a unique starting token and satisfy some subtle conditions on follow sets of base language non-terminals \cite{conf/pldi/SchwerdfegerW09}.
-However, simple starting tokens like \lstinline{pattern} cannot be guaranteed to be globally unique, so we would need to use a more verbose token like \lstinline{edu_cmu_wyvern_rx_pattern}.
-There is no simple, principled way to define scoped abbreviations for starting tokens because this mechanism is language-external.
-
-In any case, we must now decide how our newly introduced derived forms desugar to forms in our base language, which in this case requires determining which module the constructors the desugaring uses will come from.
-Clearly, simply assuming that a module named \lstinline{P} satisying \lstinline{PATTERN} is in scope is a brittle solution.
+In response to this problem, Schwerdfeger and Van Wyk have developed a modular analysis that accepts only grammar extensions that use a unique starting token and satisfy some subtle conditions on follow sets of base language non-terminals \cite{conf/pldi/SchwerdfegerW09}.
+Such extensions can be used together in any combination.
+However, simple starting tokens like \lstinline{pattern} cannot be guaranteed to be globally unique, so we would need to use a more verbose starting token like \lstinline{edu_cmu_verse_rx_pattern}.
+There is no principled way to define local abbreviations for starting tokens because this mechanism is language-external.
+
+Putting this issue aside, we must now decide how our newly introduced derived forms elaborate to forms in our base grammar.
+In this case, because we are using a modular encoding of patterns, we must first decide which module the desugaring will use.
+Clearly, simply assuming that a module identified as \lstinline{P} satisying \lstinline{PATTERN} is in scope is a brittle solution.
 Indeed, we should expect that the extension mechanism actively prevents such capture of specific variable names to ensure that variables (including module variables) can be freely renamed.
-However, we note that such \emph{hygiene mechanisms} are only well-understood when performing term-to-term rewriting (as in Lisp-style \emph{macro systems}\todo{citation}) or simple equational systems like those found in Coq.
+Such a \emph{hygiene discipline} is well-understood when performing term-to-term rewriting (discussed below) or in simple rewrite systems like those found in Coq.
 For more flexible mechanisms, the issue is a topic of ongoing research (none of the grammar-based mechanisms described above enforce hygiene).
 
-Putting aside the question of hygiene, we can address the problem by requiring that the client explicitly identify the module the desugaring should use:
-\begin{lstlisting}[numbers=none]
-let N = edu_cmu_wyvern_rx_pattern[P] /A|T|G|C/
-\end{lstlisting}
-Syntactically, this is the best we can do using existing approaches.
-
-These approaches further suffer from a paucity of direct reasoning principles, i.e.
-the program can only be reasoned about post-desugaring.
+Putting aside the question of hygiene as well, we can address the problem by requiring that the client explicitly identify the module the desugaring should use:
+\begin{lstlisting}[numbers=none]
+let ssn = edu_cmu_verse_rx_pattern P /\d\d\d-\d\d-\d\d\d\d/
+\end{lstlisting}
+For patterns constructed compositionally, we can define \emph{splicing} syntax for both strings and other patterns:
+\begin{lstlisting}[numbers=none,escapechar=|]
+fun example_syn(name : string)
+  edu_cmu_verse_rx_pattern P /@name: %ssn/
+\end{lstlisting}
+The body of this function elaborates to the body of \lstinline{example_fixed} as shown above.
+Had we mistakenly used the pattern splicing prefix, \lstinline{%name}, rather than the string splicing prefix, \lstinline{@name}, we would get a static type error, rather than a silent security vulnerability.
+
+Such syntax extension approaches suffer from a paucity of direct reasoning principles.
 Given an unfamiliar piece of syntax, there is no simple method for determining what type it will have, or even for identifying which extension determines its desugaring, causing difficulties for both humans (related to code comprehension) and tools.
+
+\subsubsection{Term Rewriting}
+An alternative approach is to leave the concrete syntax of the language fixed, but repurpose it for novel ends using a \emph{local term-rewriting system}.
+Among the earliest examples is the LISP macro system \cite{Hart63a}.
+This system was later refined, notably in Scheme and its dialects to better support reasoning compositionally about the rewriting that is performed \cite{Kohlbecker86a}.
+In languages with a stronger static type discipline, variations on macros that restrict rewriting to a particular type and perform the rewriting in an explicitly staged manner have also been studied \cite{Herman10:Theory,ganz2001macros} and integrated into languages, e.g.
+MacroML \cite{ganz2001macros} and Scala \cite{ScalaMacros2013}.
+
+The most immediate problem with using these in our example is that we are not aware of any such statically-typed macro system that integrates cleanly with an ML-style module system.
+However, let us imagine a macro system that would allow us to repurpose string syntax  as follows:
+\begin{lstlisting}[numbers=none]
+let ssn = pattern P {rx|SSTR\d\d\d-\d\d-\d\d\d\dESTR|rx}
+\end{lstlisting}
+
+Here, \lstinline{pattern} is a macro parameterized by a module \lstinline{P : PATTERN}.
+It statically parses the provided string literal (which must still be written using an Ocaml-style literal here) to generate an elaboration, at type \lstinline{P.t}.
+
+For patterns that are constructed compositionally, we need to get more creative. For example, we might repurpose the infix operators that the language provides for other purposes:
+
+\begin{lstlisting}[numbers=none,escapechar=|]
+fun example_macro(name : string)
+  pattern P (name ^ "SSTR: ESTR" + ssn)
+\end{lstlisting}
+
+This does not give us syntax that is quite as clean and clear as the syntax available using syntax extensions, but it does address a number of its deficiencies: there cannot be syntactic conflicts (because the syntax is not extended), we can define macro abbreviations because macros are integrated into the language, there is a hygiene discipline that guarantees that the elaboration will not capture variables inadvertently, and there is a typing discipline (i.e.
+we need not determine the elaboration to know what type it will have).
 
@@
 \subsection{Contributions}\label{sec:syntax-contributions}
-We propose designing a  syntax extension mechanism that addresses all of the problems just described. It consists of two constituent concepts:
-\begin{itemize}
-\item The concept that the mechanism is oriented around is the \emph{typed syntax macro} (TSM).
-%A TSM is applied to a \emph{delimited form}, which can contain  arbitrary syntax in its \emph{body}, and zero or more module or type parameters.
-Based on the body of the delimited form, the TSM statically computes an elaboration to a {base form}.
-For example, consider the following concrete term:
-\begin{lstlisting}[numbers=none]
-pattern[P] /SURLA|T|G|CEURL/
-\end{lstlisting}
-
-The TSM \lstinline{pattern} is being applied to a module parameter, \lstinline{P}, and a \emph{delimited form}, \lstinline{/SURLA|T|G|CEURL/}. This term elaborates \emph{statically} to the following:
+We propose a new primitive, the \textbf{typed syntax macro} (TSM), that combines the flexibility of syntax extensions with the reasoning guarantees of statically-typed macros, and show how to integrate it into a language with an ML-style module system, addressing all of the problems described above.
+We then introduce \textbf{type-specific languages} (TSLs), which leverage local type inference to control TSM dispatch.
+This further decreases syntactic cost (which is, of course, the entire purpose of this exercise).
+The result is that derived syntax defined using TSLs  approaches the cost of derived syntax built in primitively.
+
+\subsubsection{Typed Syntax Macros (TSMs)}
+%A typed syntax macro is invoked by applying it to a \emph{delimited form}, which can contain  arbitrary syntax in its \emph{body}.  
+As our first example, consider the following concrete term:
+\begin{lstlisting}[numbers=none]
+pattern P /SURLA|T|G|CEURL/
+\end{lstlisting}
+Here, a parameterized TSM \lstinline{pattern} is applied to a module parameter, \lstinline{P}, and a {delimited form}, \lstinline{/SURLA|T|G|CEURL/}.
+Based on the body of the delimited form, i.e.
+the characters between the delimiters, the TSM statically computes an elaboration to the following {base external form}:
 
@@
 \end{lstlisting}
-
-The TSM is defined as shown below:
-\begin{lstlisting}[numbers=none]
-syntax pattern[P : PATTERN] for P.t
+The parameterized TSM \lstinline{pattern} is defined as shown below:
+\begin{lstlisting}[numbers=none]
+syntax pattern(P : PATTERN) at P.t
   fn (ps : ParseStream) => (* pattern parser here *)
@@
 We declare a module parameter (calling it \lstinline{P} here is unimportant, i.e.
-the TSM can be used with \emph{any} module satisfying module type \lstinline{PATTERN}), and the type clause \lstinline{for P.t}, which guarantees that any use of this TSM will either be ill-typed or elaborate to a term of type \lstinline{P.t}.
-The elaboration is computed by the compile-time action of the parse function defined in the indented block above.
-This function must be of type \lstinline{ParseStream -> Exp}, where the type \lstinline{ParseStream} gives the function access to the \emph{body} of the delimited form (in blue above) and the type \lstinline{Exp}  encodes the abstract syntax of the base language.
-Both types are defined in the Wyvern prelude, which is a set of definitions available ambiently.
-
-When the type of the term is known, e.g. due to a type annotation on the let binding, the module parameter \lstinline{P} can be inferred:
+the TSM can be used with \emph{any} module satisfying module type \lstinline{PATTERN}), which is bound in the scope of the type clause \lstinline{at P.t}.
+This clause  guarantees that all elaborations successfully generated by this TSM will be of type \lstinline{P.t}.
+Elaborations are generated by the static action of the parse function defined in the indented block on the next line.
+The parse function must be of type \lstinline{ParseStream -> Exp}, where the type \lstinline{ParseStream} gives the function access to the \emph{body} of the delimited form (in blue above) and the type \lstinline{Exp} is a case type that encodes the abstract syntax of the base language (i.e.
+the Verse EL).
+Both types are defined in the Verse \emph{prelude}, which is a set of definitions available ambiently.
+
+In positions where the type is known, e.g. due to a type ascription or at the argument position of a function call, the module parameter \lstinline{P} can be inferred:
 \begin{lstlisting}[numbers=none]
@@
 
-TSMs can be abbreviated using a let-binding style mechanism:
-\begin{lstlisting}[numbers=none]
-let syntax pat = pattern[P]
-pat /SURLA|T|G|CEURL/
-\end{lstlisting}
-
-TSMs can parse portions of the body of the delimited form as a base language term, to support splicing syntax:
-\begin{lstlisting}[numbers=none]
-let N = pat /SURLA|T|G|CEURL/
-let BisI = pat /SURLGC({EURLNSURL})GCEURL/
-\end{lstlisting}
-A hygiene mechanism ensures that only those portions of the elaboration derived from such spliced terms can refer to variables in the surrounding scope.
-
-Strings are never involved, so the attendant security issues described previously are easily avoidable. That is, the following
-
-\item To support an even more concise usage profile, Wyvern also  supports \emph{type-specific languages} (TSLs), which allow library providers to associate a TSM directly with a declared type.
-For example, the module \lstinline{P} above can associate \lstinline{pattern} with \lstinline{P.t}.
-Local type inference then determines which TSM is applied implicitly to handle a form not prefixed by a TSM name.
-For example, the following is equivalent to the above:
-\begin{lstlisting}[numbers=none]
-let N : P.t = /SURLA|T|G|CEURL/
-\end{lstlisting}
-\end{itemize}
-\subsubsection{Typed Syntax Macros (TSMs)}\label{sec:tsms}
-For simplicitly, let us begin with a typed syntax macro providing pattern syntax using the case type encoding above:
-\begin{lstlisting}[numbers=none]
-syntax pattern => Pattern = e_parser
-\end{lstlisting}
-The term \lstinline{e_parser}, elided here, must have type \lstinline{ParseStream -> Exp}, where \lstinline{ParseStream} classifies a sequence of characters and \lstinline{Exp} classifies a reified Wyvern expression.
-
-\subsubsection{Type-Specific Languages (TSLs)}\label{sec:tsls}
+TSMs can be partially applied and abbreviated using a let-binding style mechanism:
+\begin{lstlisting}[numbers=none]
+let syntax pat = pattern P 
+let ssn = pat /SURL\d\d\d-\d\d-\d\d\d\dEURL/
+\end{lstlisting}
+
+TSMs can treat portions of the body of the delimited form as a spliced base language term, to support splicing syntax:
+\begin{lstlisting}[numbers=none]
+fun example_tsm(name: string)
+  pat /SURL@EURLnameSURL: %EURLssn/
+\end{lstlisting}
+A hygiene mechanism ensures that only those portions of the elaboration derived from such spliced terms can refer to variables in the surrounding scope, preventing inadvertent variable capture by the elaboration.
+In other words, the generated elaboration must be context-independent.
+
+\paragraph{Formalization} To give a flavor of the formal specification underlying TSMs, let us look at the rule for handling TSM invocation in a synthetic position. 
+\begin{mathpar}
+\inferrule[syn-aptsm]{
+    \Delta \vdash s ~@~ \sigma~\{\iota_\text{parser}\}\\
+    \mathbf{parsestream}(body) = \iota_\text{ps}\\
+    \iota_\text{parser}(\iota_\text{ps}) \Downarrow \iota_\text{elab}\\
+    \iota_\text{elab} \uparrow e_\text{elab}\\\\
+    \Delta; \Gamma; \emptyset; \emptyset \vdash e_\text{elab} \Leftarrow \sigma \leadsto \iota
+}{
+    \Delta_\text{out}; \Gamma_\text{out}; \Delta; \Gamma \vdash \mathbf{invoketsm}(s; body) \Rightarrow \sigma \leadsto \iota
+}
+\end{mathpar}
+Note that the judgement forms are  slightly simplified in that they omit various contexts relevant only to other Verse primitives.
+We include only \emph{typing contexts}, $\Gamma$, which map variables to types, and \emph{type abstraction contexts}, $\Delta$, which track universally quantified type variables (both in the standard way).
+The premises have the following purposes:
+\begin{enumerate}
+\item The first premise can be pronounced ``Under $\Delta$, TSM expression $s$ is a TSM at type $\sigma$ with parser $\iota_\text{parser}$''.
+Application of a parameterized TSM to its parameters is handled by this judgement (not shown).
+\item The second premise creates a parse stream, $\iota_\text{ps}$, from the body of the delimited form.
+\item The third premise applies the parser to the parse stream to generate an encoding of the elaboration, $\iota_\text{elab}$.
+\item The fourth premise decodes $\iota_\text{elab}$, producing the elaboration $e_\text{elab}$.
+\item The fifth premise validates the elaboration by analyzing it against the type $\sigma$ under an empty context.
+The \emph{current} typing and kinding contexts are ``saved'' for use when a spliced term is encountered during this process by setting them as the new \emph{outer contexts}.
+
+Variables in the outer contexts cannot be used directly by the elaboration, e.g. they are ignored by the (syn-var) rule:
+\begin{mathpar}
+\inferrule[syn-var]{ }{
+    \Delta_\text{out}; \Gamma_\text{out}; \Delta; \Gamma, x : \sigma \vdash x \Rightarrow \sigma \leadsto x
+}
+\end{mathpar}
+The outer contexts are switched back in only when encountering a spliced expression, which is marked:
+\begin{mathpar}
+\inferrule[syn-spliced]{
+    \emptyset; \emptyset; \Delta_\text{out}; \Gamma_\text{out} \vdash e \Rightarrow \sigma \leadsto \iota
+}{
+    \Delta_\text{out}; \Gamma_\text{out}; \Delta; \Gamma \vdash \mathbf{spliced}(e) \Rightarrow \sigma \leadsto \iota
+}
+
+\inferrule[ana-spliced]{
+    \emptyset; \emptyset; \Delta_\text{out}; \Gamma_\text{out} \vdash e \Leftarrow \sigma \leadsto \iota
+}{
+    \Delta_\text{out}; \Gamma_\text{out}; \Delta; \Gamma \vdash \mathbf{spliced}(e) \Leftarrow \sigma \leadsto \iota
+}
+\end{mathpar}
+For example, the elaboration generated in \lstinline{example_tsm} above would be:
+\begin{lstlisting}[numbers=none]
+P.Seq(P.Str(spliced(name)), P.Seq(P.Str ": ", spliced(ssn)))
+\end{lstlisting}
+\end{enumerate}
+
+We will give a more complete account of all of these judgements later.
+\subsubsection{Type-Specific Languages (TSLs)}
+To further lower the syntactic cost of using TSMs, Verse also  supports \emph{type-specific languages} (TSLs), which allow library providers to associate a TSM directly with a declared type.
+For example, a module \lstinline{P} can associate \lstinline{pattern} with \lstinline{P.t} as follows:
+\begin{lstlisting}[numbers=none]
+module P : PATTERN with syntax pattern at t
+  type t = (* ... *)
+  (* ... *)
+\end{lstlisting}
+
+ Local type inference then determines which TSM is applied when analyzing a delimited form not prefixed by a TSM name. For example, the following is equivalent to the above:
+\begin{lstlisting}[numbers=none]
+fun example_tsl(name : string) : P.t
+  /SURL@EURLnameSURL: %EURLssn/
+\end{lstlisting}
+
+As another example, we can use TSLs to express derived list syntax. For example, if we use a case type, we can declare the TSL directly upon declaration as follows:
+\begin{lstlisting}[numbers=none]
+casetype list('a) {
+  Nil
+  Cons of 'a * list('a)
+} with syntax
+  fn (body : ParseStream) => (* ... comma-delimited spliced exps ... *)
+\end{lstlisting}
+Together, this allows us to write a list of patterns like this:
+\begin{lstlisting}[numbers=none]
+let x : list(P.t) = [/SURL\dEURL/SHTML, EHTML/SURL\d\dEURL/SHTML, EHTML/SURL\d\d\dEURL/]
+\end{lstlisting}
 
 \subsection{Timeline}\label{sec:syntax-timeline}
-SAC, ECOOP.
-
-TODO: module-parameterized TSMs
-
-\section{Extensible Semantics}\label{sec:extensible-semantics}
-
-\subsubsection{Example 2: Regular Strings}\label{sec:rstr}
-The example above deals with how regular expression patterns are  encoded, introduced and reasoned about.
+We have described and formally specified TSMs in a recently published paper \cite{sac15}, and TSLs in a paper last year \cite{TSLs}, both in the context of the Wyvern language.
+In the context of Verse, there are some changes that need to be made in the formalization.
+In particular, neither paper covered integration with an ML-style module system.
+We plan to complete this in the course of writing the dissertation.
+
+% \subsection{Typed Syntax Macros (and Type-Specific Languages)}
+% Consider the following \emph{recursive case type}, which encodes the tree structure of HTML documents (we will see how case types are themselves library constructs later):
+% \begin{lstlisting}[numbers=none]
+% casetype HTML
+%   Empty
+%   Seq of HTML * HTML
+%   TextNode of string
+%   H1Element of Attributes * HTML (* Attributes not shown *)
+%   (* ... *)
+% \end{lstlisting}
+% We can introduce values of this type by applying one of these declared constructors to data of the corresponding type.
+For example, here is a function that creates a heading given a string (assuming the variable \lstinline{empty_attrs : Attributes} is in scope, not shown):
+% \begin{lstlisting}[numbers=none]
+% fun make_heading(x : string) -> HTML
+%   H1Element(empty_attrs, TextNode(x))
+% \end{lstlisting}
+% The same function can be written using more idiomatic HTML syntax via a TSM as follows:
+% \begin{lstlisting}[numbers=none]
+% fun make_heading_tsm(x : string)
+%   html `SHTML<h1><%EHTMLxSHTML></h1>EHTML`
+% \end{lstlisting}
+% Here, \lstinline{html} identifies a TSM defined as below:
+% \begin{lstlisting}[numbers=none]
+% syntax html for HTML
+%   fn (body : ParseStream) -> Exp => (* ... *)
+% \end{lstlisting}
+% The function in the indented block is invoked {statically} to generate an \emph{elaboration} based on the {body} of the \emph{delimited form} provided to the TSM, here \lstinline{SHTML<h1><%EHTMLxSHTML></h1>EHTML},
+% for subsequent typechecking and translation to the Verse IL.
+In this example, the elaboration generated is equivalent to the one shown above in \lstinline{make_heading}.
+Both \lstinline{ParseStream}, which provides access to the body as a stream of characters, and \lstinline{Exp}, which encodes the abstract syntax of the Verse EL, are types defined in the Verse \emph{prelude}, which is simply a collection of definitions loaded before all others.
+The implementation of this parse function is not particularly interesting, so we omit it here.
+We will discuss how generating this function from a declarative grammar specification is itself a mode of use of TSMs later.
+
+% Notice that a portion of the body, indicated in black for clarity, is treated as a \emph{spliced expression}.
+The TSM determines what syntactically distinguishes spliced expressions (here, the delimiters \lstinline{<%} and \lstinline{>} around an identifier or, not shown, \lstinline{<%>} and \lstinline{</%>} around an expression).
+% Verse only specifies the delimiters that can be used to transition from the expression language to the TSM body (here, we used backticks, but Verse supports several others, including \emph{layout delimiters}, which we will discuss later).
+This ensures that syntactic conflicts cannot arise by construction and localizes reasoning about syntactic ambiguity to each TSM, without incurring substantial syntactic cost.
+
+% The semantics includes an \emph{elaboration validation} step that maintains a \emph{hygiene discipline} by ensuring that the elaboration does not make any assumptions about the context around the TSM application.
+In particular, only portions of the elaboration derived from spliced expressions can refer to variables in the surrounding context.
+Moreover, by requiring that the TSM provide a type annotation, it maintains a \emph{typing discipline}.
+For example, due to the type annotation on the definition of \lstinline{html} above, clients can assume that all successful applications of the TSM will elaborate to a term of type \lstinline{HTML}, without needing to reason about the elaboration logic itself.
+Not shown here, this annotation can be parameterized by types as well as modules.
+Because the Verse module system is borrowed directly from ML,  this means that TSMs can be used to define derived syntax valid for all instances of an abstract type.
+We will discuss this after motivating it with another example in Section \ref{sec:syntax}.
+
+% To decrease the syntactic cost of applying a TSM even further (controlling syntactic cost being, of course, the very purpose of TSMs), Verse can often locally infer these parameters.
+Moreover, when a type is declared, Verse allows the library provider to designate one TSM as its \emph{type-specific language} (TSL).
+Clients can then omit mention of this TSM entirely in positions where an expression of that type is expected.
+For example, if \lstinline{html} was designated as the TSL for \lstinline{HTML}, then we could have written the function above equivalently as:
+% \begin{lstlisting}[numbers=none]
+% fun make_heading_tsl(x : string) -> HTML
+%   'SHTML<h1><%EHTMLxSHTML%></h1>EHTML'
+% \end{lstlisting}
+
+%These mechanisms are one reason why we specify Verse by a bidirectionally typed elaboration semantics.
+
+\section{Modularly Metaprogrammable Type Structure}\label{sec:metamodules}
+The Verse EL also does not build in substantial type structure.
+Only functions are built in directly.
+The type structure of case types (shown in use above), tuples, records, objects and other more esoteric constructs that we will describe later can all be expressed as modes of use of \emph{metamodules} (which are distinct from, though conceptually related to, \emph{modules}).
+
+The remainder of this section is organized like Section \ref{sec:syntax}.
+We begin with an example of type structure that cannot be expressed in languages like ML directly in Sec.
+\ref{sec:metamodules-example}, then show how existing approaches are insufficient in Sec.
+\ref{sec:metamodules-related}.
+Then we outline our approach in Sec.
+\ref{sec:metamodules-solution} and conclude with a timeline for remaining work in Sec.
+\ref{sec:metamodules-timeline}.
+
+\subsection{Motivating Example: Labeled Products and Regular Strings}\label{sec:metamodules-example}
+As a simple introductory example, let us consider \emph{labeled products} and \emph{regular strings}. In Verse, a simple labeled product type classifying conference papers can be defined like this:
+\begin{lstlisting}[numbers=none]
+let type Paper = lprodtype {SPCT
+  title : EPCTrstring /SURL.+EURL/
+   SPCTconf : EPCTrstring /SURL([A-Z]+) (\d\d\d\d)EURL/
+}
+\end{lstlisting}
+The first row is labeled \lstinline{title} and specifies the regular string type \lstinline{rstring /SURL.*EURL/}.
+Regular string types classify values that behave as strings known to be in the regular language specified by the regular expression given as the type parameter.
+Here, we have that paper titles must be non-empty strings.
+The second row is labeled \lstinline{conf} and specifies a regular string type with two parenthesized groups, corresponding to the conference abbreviation and the year.
+The \lstinline{let type} construction defines a type synonym, \lstinline{Paper}. Alternatively, we could have used the following construction to declare \lstinline{Paper} \emph{generatively}:
+\begin{lstlisting}[numbers=none]
+lprodtype Paper {
+  SPCTtitle : EPCTrstring /SURL.+EURL/
+   SPCTconf : EPCTrstring /SURL([A-Z]+) (\d\d\d\d)EURL/
+}
+\end{lstlisting}
+The difference is in how type equality is handled. Two labeled product types with the same signature are equal, unless declared generatively, in which case type equality is nominal.
+
+In either case, we can introduce a value of type \lstinline{Paper} in an analytic position in one of two ways. We can omit the labels:
+\begin{lstlisting}[numbers=none]
+let exmpl : Paper = {"SUSAn Example PaperEUS"SCSS, ECSS"SUSEXMPL 2015EUS"}
+\end{lstlisting}
+Alternatively, we can include them explicitly for clarity:
+\begin{lstlisting}[numbers=none]
+let exmpl : Paper = {SCSStitle => ECSS"SUSAn Example PaperEUS"SCSS, conf => ECSS"SUSEXMPL 2015EUS"}
+\end{lstlisting}
+
+We can then project a row out of such a value using the row projection operator:
+\begin{lstlisting}[numbers=none]
+let exmpl_conf = # `SNATconfENAT exmpl 
+\end{lstlisting}
+Here, \lstinline{#} is an operator parameterized by a static label, written literally \lstinline{`SNATconfENAT}, and taking one argument, \lstinline{exmpl}.
+
+We can also project captured groups out of a regular string using the \lstinline{#group} operator, which is parameterized by a group index:
+\begin{lstlisting}[numbers=none]
+let exmpl_conf_name = #group 0 exmpl_conf
+\end{lstlisting}
+
+The variable \lstinline{exmpl_conf_name} has type \lstinline{rstring /SURL[A-Z]+EURL/}.
+
+Both labeled products and regular strings support a number of other operations. For example, labeled products can be concatenated (with any common rows updated with the value on the right):
+\begin{lstlisting}[numbers=none]
+let a : lprodtype {SPCTauthors : EPCTlist(rstring /SURL.+EURL/)} = {["SUSHarry Q. BovikEUS"]}
+let exmpl_paper_final = lprod+ exmpl_paper a
+\end{lstlisting}
+We can drop a row as well:
+\begin{lstlisting}[numbers=none]
+let exmpl_paper_anon = lprod- exmpl_paper_final `SNATauthorsENAT
+\end{lstlisting}
+
+\subsection{Existing Approaches}\label{sec:metamodules-related}
+\subsubsection{Records and Tuples}
+Recall that Verse builds in only nullary and binary products into its internal language, so the type structure described above for labeled products  is not directly expressible using IL primitives.
+If we weaken slightly our insistence on minimalism and built in record/tuple types, as in other functional languages, we have  options that get close to (but do not  exactly achieve) this type structure.
+In particular, unlike labeled product types, record types are identified up to row reordering.
+Using a record with the same row labels would prevent us from omitting the labels and relying on positional information when introducing a value of record type in an analytic position, as we did with type \lstinline{Paper} above.
+%If we instead use an unlabeled product type (which in ML is actually a record type with numeric labels, which are omitted syntactically), we lose the ability to use labels in the elimination form.
+
+One workaround is to replace each labeled product type in the program with a record type with row labels tagged by position (recovering the type disequality we seek to express with labeled product types) and conversion functions from the corresponding unlabeled product type and untagged record type to this type.
+This partially recovers some of the syntactic cost we seek, but introduces much additional  boilerplate as well as run-time cost.
+We are interested in such issues, so this is not an entirely satisfying solution.
+% In any case, records, tuples and labeled tuples are all library constructs in Verse.
+Moreover, this still does not give us support for the extra operations, like concatenation,  shown above.
+
+\subsubsection{Refinement Types}
+For regular strings, one could instead use standard strings (which themselves may be lists or vectors of characters together with derived syntax), inserting run-time checks around introductory forms and into other operations (with one exception, discussed below).
+This would, of course,  weaken the static semantics and add both syntactic and run-time cost.
+To recover the static guarantees of regular strings, we could treat membership of a string in a specified regular language as a static verification condition, moving the logic governing regular string types out of the language and into an external \emph{refinement type system} \cite{Freeman91}.
+
+One problem with this approach is that there is no direct analog to the group projection operator described above.
+The static and dynamic semantics of this operator depend on information available only in the static type of the regular string (i.e.
+the position of the captured group).
+Refinement types are strictly a layer above the language, and thus cannot affect its dynamic semantics.
+
+A related problem is that there is no feasible way to use refinement types to introduce type disequalities that can later affect the internal representation of a value.
+All standard strings are represented uniformly.
+The fact that some value of type \lstinline{string} can be given a refined type \lstinline{rstring /A|B/} cannot be justification to locally alter its representation (e.g.
+to a single bit in this case).
+For regular strings, there may be limited benefit to doing this, but in general, this can have significant performance ramifications.
+For example, it is common to justify the omission of a primitive type of ``unsigned integers representable in 32 bits'' by the fact that this is ``merely'' a refinement of arbitrary precision integers, but in so doing one loses the ability to use a machine representation more suitable to this invariant.
+%Refinement types cannot be used for this purpose without the possibility of conflict.
+This is also the reason why one cannot viably do away with primitive record types by claiming that they are refinements of finite mappings.
+
+Put another way, refinement types are useful when one only needs to make finer static distinctions within an existing type.
+When one needs to introduce new operators or type disequalities that have ``downstream'' impact, refinement types are not the right solution.%Refinement types are useful when no non-trivial operators need to be introduced into the language and the refinement has no impact on representation, but the type system itself must be extended in some way in more advanced situations.
+
+\subsection{Contributions}\label{sec:metamodules-approach}
+
+An important point is that labeled products, as well as other ``record-like'' constructs, can  be expressed by a \emph{type-directed translation} targeting our IL (and indeed, many compilers take advantage of facts like this to simplify the number of cases that must be considered downstream).
+%it is inaccurate to state that they are merely syntactic sugar over such a language.
+%Object systems are perhaps even more diverse across languages\todo{citation}.
+%Other dialects are justified by a desire to provide more special-purpose semantic constructs. For example, 
+Substantially more sophisticated examples require at most only a very slightly more capable intermediate language (the IL of popular virtual machines, even those with flaws like ubiquitous \texttt{null} references, e.g.
+JVM or CIL bytecode, are also sufficient if used strictly as translation targets).
+
+Metamodules introduce type and operator constructors parameterized by arbitrary \emph{static values} and metaprogrammatically express the logic governing their typechecking and  translation to the IL using the Verse \emph{static language} (SL).
+Metamodules are classified by \emph{metasignatures}.
+For example, let us consider the following metasignature:
+\begin{lstlisting}
+metasignature LPROD
+  tycon c of LblTyList
+  ana opcon intro_unlabeled
+  ana opcon intro_labeled of LblList
+  syn opcon # of Lbl
+\end{lstlisting}\todo{reworking active tycons as metasignatures and metastructures is compelling, but I need to figure it out. will continue writing.}
+
+% The semantics uses internal type abstraction to guarantee that metamodule providers can reason modularly about translation invariants, just as it uses external type abstraction to guarantee that module providers can reason modularly about representation invariants.%For example, a metamodule that expresses the type structure of records would introduce a type constructor  \texttt{record} parameterized by a static mapping from labels to types, and two operators: the introductory operator \texttt{intro}, parameterized by a list of row labels, and the row projection operator \texttt{prj}, parameterized by a row label.
+It might then choose to internally represent records by binary products, which the IL provides.
+Critically, the choice of internal representation is \emph{computed} based on each record type's parameter, and it is held abstract outside the metamodule using internal type abstraction, facilitating modular reasoning about translation invariants (just as external type abstraction facilitates modular reasoning about representation invariants in the module system).
+Metamodules like this can express many other constructs, including variants of records that support additional operations, labeled sums, object systems, constrained strings, typechecked operations on format strings like \texttt{printf} and typechecked foreign function interfaces.
+By using %This simplifies the Verse core language relative to the core languages of ML,  Scala and other comparable languages.
+
+% These constructs, very generally summarized, give library providers \emph{metaprogrammatic control} over aspects of the \emph{external} (i.e.
+user-facing) textual syntax and type structure of Verse, respectively.
+This \emph{external language} (EL) is given meaning by type-directed translation to a minimal \emph{typed internal language} (IL), specified in a completely standard manner.
+In some ways, this style of language specification is comparable to a specification for the first stage of a type-directed compiler shifted ``one level up'' into the language and modularly reorganized.
+By \emph{metaprogrammatic control}, we mean that Verse exposes a \emph{static language} (SL) to library providers.
+Functions written in the SL, which we refer to as \emph{static functions}, are invoked by the external semantics to control certain aspects of parsing, elaboration, typechecking and translation.
+%This IL can be specified in a completely standard manner, but the user-facing \emph{external language} (EL) must be specified in a novel manner to support these constructs.
+
+%  %while imposing constraints in order to maintain strong metatheoretic guarantees, including the guarantee that these constructs can be reasoned about modularly and used together by clients in any combination, without the possibility of syntactic or semantic conflicts.\footnote{...other than simple naming conflicts, which must be handled by adopting appropriate naming conventions, e.g.
+based on URIs as in the Java ecosystem.} By \emph{metaprogrammatic}, we mean that Verse exposes a \emph{static language} (SL) to library providers.
+Functions written in the SL, which we refer to as \emph{static functions}, are invoked by the language to control certain aspects of parsing, elaboration, typechecking and translation to Verse's minimal typed internal language (IL), which is specified in the standard manner.% In a sense, Verse pulls machinery that is usually part of a compiler into the language specification so that it can give library providers greater control over it.
+
+% %Although metamodules are in some sense conceptually prior to TSMs, it will be  pedagogically useful for us to introduce TSMs first, in Sec.
+\ref{sec:syntax}, assuming temporarily that the core type structure is specified in the standard manner.
+We then show how metamodules can influence the type structure in Sec.
+\ref{sec:metamodules}.
+
+% %This trend toward languages that are a growing hodge podge of constructs surrounded by an ecosystem of ``toy dialects'' can only be slowed by the development of mechanisms that are expressive enough to subsume large classes of existing constructs, pulling them out of the language definition and into modularly composable libraries.
+
+% % The ability to reason separately about libraries and import them freely is critical when constructing large software systems and participating in modern software ecosystems.
+If using a library requires a developer to directly modify the programming language they are using, detecting and resolving conflicts between constructs manually, then reestablish critical metatheoretic properties about the language followed by important invariants throughout their own code, the library will be essentially useless to them.
+% % %This means that it is difficult to construct large software systems and ecosystems from components written in different dialects. 
+% % From this perspective, it is hardly surprising when software engineers avoid dialects and settle on a single common language, where they can rely on the guarantees provided by its module system to program ``in the large''.
+%The designers of these languages may slowly adopt the most generally useful new ideas.
+% % Unfortunately, this has left many useful constructs for programming ``in the small''  languishing in ``toy dialects''.% It is safe to say that in contemporary times, language dialects should be considered harmful.
+% It is little wonder that many programmers eschew esoteric language dialects in favor of languages that stopped evolving decades ago.
+
+% % It is nearly always the case, however, that the dynamic semantics of these dialects can be defined by type-directed translation to the existing language.
+For example, ReactiveML builds in support for functional reactive programming, but its dynamic semantics can be implemented by type-directed translation to Standard ML \cite{mandel2005reactiveml}.
+In some cases, the semantics of the dialect are directly specified in this way.
+For example, F\# \cite{syme2012expert} introduces functional datatypes and the other constructs of the ML core language, defining them in terms of constructs available in the Common Intermediate Language (CIL), a class-based object-oriented language.
+Ocaml includes a class-based object system that can be defined by type-directed translation to Caml, a language building in little more than recursive product and sum types.
+
+% %{\Verse} aims to explore the polar opposite design philosophy: a core language designed to remain small and as unopinionated as possible.
+Rather than building in extensive syntax and type structure \emph{a priori}, it is organized around metaprogramming mechanisms that  %In other words, power is decentralized (though not completely, as we will discuss), while responsibility over metatheoretic guarantees remains ultimately in the hands of the language designer.
+
+% Ensuring that the language specification is  metatheoretically well-behaved and, in particular, that the constructs that we introduce can be reasoned about modularly and used together in any combination is the primary challenge in this work.
+% For example, if we simply conceived of the language as being specified by a ``bag of forms and rules'' (like ``paper'' specifications of languages often appear) and let library providers introduce new forms and rules, as much of the related work does, one could easily cause syntactic conflicts, or introduce rules that violate type safety, interfere with invariants being maintained by rules defined in other libraries, or introduce undesirable forms of non-determinism into the semantics.
+%There is also the question of how to allocate finite syntactic resources.
+For example, as noted above, the design space around record types is large.
+If we allow libraries to explore this design space, which library gets to determine the meaning of a form like \lstinlinep{\{label1=e1, label2=e2\}}?
+
+%This suggests a need for language-integrated mechanisms that allow library providers to modularly express syntactic and semantic constructs of the sort that today require new dialects, allowing for perhaps some low-cost syntactic compromises.
+%For example, it should be possible to express different record systems within a language specified like ML (or the simply typed lambda calculus).
+The challenge  is to give formal shape to these intuitions.
+%Our broad aim in this thesis is to design language mechanisms that take significant steps toward achieving this goal.
+
+% \section{Contributions}
+% \subsection{Verse}
+% To provide a coherent vehicle for our contributions, we introduce a new programming language called Verse.
+Let us briefly review its organization.
+Syntactically, Verse has a layout-sensitive textual concrete syntax (i.e.
+newlines and indentation are meaningful).
+This choice is not fundamental to our proposed mechanisms, but it will be useful for cleanly expressing a class of examples that we will discuss later.
+Semantically, Verse is organized into a \emph{module language} and a \emph{core language}.
+The module language is based directly on the Standard ML module language, with which we assume familiarity for the purposes of this proposal \cite{harper1997programming}.
+We will use it in an example in Section \ref{sec:syntax}.
+The core language is split into a user-facing \emph{external language} (EL) and a minimal \emph{typed internal language} (IL).
+The EL  is specified by a type-directed translation semantics targeting the much simpler IL.
+Notionally, this can be thought of as simply shifting the first stage of a type-directed compiler directly into the language specification \cite{tarditi+:til-OLD}.
+More specifically, we specify a {bidirectionally typed translation semantics}, i.e.
+one where a distinction is made between \emph{type analysis} (the type is an ``input'') and \emph{type synthesis} (the type is an ``output''), for reasons that we will return to later.
+
+% The main novelty is that the Verse EL builds in almost no type structure or syntax \emph{a priori}.
+Instead, \textbf{metamodules} and \textbf{typed syntax macros} (TSMs), described below, are expressive enough to render building in substantial type structure and concrete syntax unnecessary.
+Although metamodules are in some sense prior to TSMs, we will describe TSMs first, temporarily assuming that the EL does build in standard type structure, for pedagogical reasons.
+
+% The mechanisms we introduce are relatively insensitive to the particular choice of IL.
+Choosing the semantics of the IL is the primary decision that a language designer leveraging these mechanisms would need to make.
+The only strict requirements are that the IL is type safe and supports parametric type abstraction.
+For our purposes, we will keep things simple by using the strictly evaluated polymorphic lambda calculus with nullary and binary product and sum types and recursive types, which we assume the reader has familiarity with and for which all the necessary metatheorems are already well-established \cite{pfpl}.
+Features like state, exceptions, concurrency primitives, scalars, arrays and other constructs characteristic of a first-stage compiler IL would also be included in practice, but we omit them when their inclusion would not meaningfully affect our discussion.
+
+% \subsection{Typed Syntax Macros}
+% Typed syntax macros (TSMs) allow library providers  to metaprogrammatically control the parsing and elaboration of delimited segments of concrete syntax.
+As such, Verse does not need to build in  many forms of derived syntax, e.g.
+list syntax as built in to many languages, but also HTML syntax, regular expression pattern  syntax and quasiquotations.
+
+% TSMs are well-behaved: syntactic conflicts cannot arise by construction and the semantics includes an elaboration validation step that maintains hygiene  and, by requiring that TSM providers provide a type annotation, a typing discipline.
+This annotation can be parameterized by types as well as modules.
+Because the Verse module system is borrowed directly from ML,  this means that TSMs can be used to define derived syntax valid for all instances of an abstract type.
+
+% To control the syntactic cost of invoking a TSM (controlling syntactic cost being, of course, the very purpose of TSMs), Verse can often locally infer these parameters.
+Moreover, when a type is declared, Verse allows the library provider to designate one TSM as its \emph{type-specific language} (TSL).
+Clients can then omit mention of this TSM entirely in positions where an expression of that type is expected.
+These mechanisms are one reason why we specify Verse by a bidirectionally typed elaboration semantics.
+
+% \subsection{Metamodules}
+% Metamodules introduce type and operator constructors parameterized by arbitrary \emph{static values} and metaprogrammatically express the logic governing their typechecking and  translation to the IL using the \emph{static metalanguage} (SL).
+This can be used to express the type structure of a wide variety of constructs, including tuples/records, labeled sums, object systems, regular strings, typechecked operations on format strings like \texttt{printf} and typechecked foreign function interfaces.
+The semantics uses internal type abstraction to guarantee that metamodule providers can reason modularly about translation invariants, just as it uses external type abstraction to guarantee that module providers can reason modularly about representation invariants.%For example, a metamodule that expresses the type structure of records would introduce a type constructor  \texttt{record} parameterized by a static mapping from labels to types, and two operators: the introductory operator \texttt{intro}, parameterized by a list of row labels, and the row projection operator \texttt{prj}, parameterized by a row label.
+It might then choose to internally represent records by binary products, which the IL provides.
+Critically, the choice of internal representation is \emph{computed} based on each record type's parameter, and it is held abstract outside the metamodule using internal type abstraction, facilitating modular reasoning about translation invariants (just as external type abstraction facilitates modular reasoning about representation invariants in the module system).
+Metamodules like this can express many other constructs, including variants of records that support additional operations, labeled sums, object systems, constrained strings, typechecked operations on format strings like \texttt{printf} and typechecked foreign function interfaces.
+By using %This simplifies the Verse core language relative to the core languages of ML,  Scala and other comparable languages.
+
+% Typed syntax macros then allow programmers to metaprogrammatically control the parsing and elaboration (i.e.
+to the EL) of delimited segments of concrete syntax at a specified type or, perhaps most interestingly, a module-parameterized family of types.
+Consequently, Verse does not need to build in derived syntax for  constructs like lists, HTML documents, regular expressions and quasiquotations.
+Syntactic ambiguities are not possible by construction and, by validating the code generated by the SL, we maintain a hygienic type discipline.
+
+% \item \textbf{Metamodules} allow library providers to express new type constructors (tycons) together with associated operators, controlling their typechecking and translation to a minimal internal language (IL) metaprogrammatically.
+As such, Verse does not need to build in constructs like tuples/records (and variants thereof), labeled sums, objects, typechecked \texttt{printf} and typechecked foreign function interfaces.
+%To support tycon modules, the core language is specified like the first stage of a compiler, i.e.
+as an external language (EL) given meaning by type-directed translation to a minimal typed internal language (IL).
+% The semantics delegates control over aspects of parsing, typechecking and translation to
+% %Metafunctions are written in a typed static language (SL).% associated with TSMs and tycon modules.
+
+% Tycon modules are also well-behaved.
+Syntactic and semantic ambiguities cannot arise by construction.
+The semantics, which is specified like the first stage of a type-directed compiler shifted up into the language, includes a translation validation step that maintains several strong metatheoretic guarantees, including type safety, hygiene and a powerful modular reasoning principle that we call \emph{conservativity}: a broad class of metatheorems relating to tycon modules need only be established in a ``closed world'', i.e.
+without needing to consider  logic expressed by any other tycon modules.
+Type abstraction plays a fundamental  role in ensuring that these metatheorems will be conserved in the ``open world'' by maintaining \emph{translation independence} between tycon modules.
+This is, encouragingly, analagous to the  in maintaining representation independence between standard modules).
+% In particular,
+% \end{enumerate}
+
+% \subsection{Modularly Metaprogrammable Concrete Syntax}
+% \emph{Typed syntax macros} (TSMs) give library providers the ability to metaprogrammatically control the parsing and elaboration of delimited segments of arbitrary concrete syntax.
+For example, list syntax, HTML syntax and regular expression syntax can all be expressed using TSMs.
+Syntactic conflicts cannot arise by construction and the semantics performs a form of translation validation that maintains a hygiene discipline as well as a typing discipline, the latter by requiring that every TSM definition specify a type annotation.
+This annotation can be parameterized by types as well as modules, providing a clean method for associating derived syntax with all instances of an abstract type.
+To control the syntactic cost of invoking a TSM (controlling syntactic cost being, of course, their very purpose), Verse can often locally infer these parameters.
+Moreover, when a type is declared, Verse allows the library provider to designate one TSM as its \emph{type-specific language} (TSL).
+Clients can then omit mention of this TSM entirely in positions where an expression of that type is expected, further decreasing syntactic cost.
+%We also briefly  introduce a related mechanism that allows for the use of non-textual interactions in introducing a value of a type.
+%Local type inference to control the invocation of the TSL, rather than having to invoke it explicitly.
+This further decreases their syntactic cost.% the same usage profile as syntax that is built in to other languages.
+
+% %Verse provides forms of type generativity \emph{Type-specific languages} (TSLs) 
+% % clean integration with the module system.
+
+% \emph{Tycon structures} allow library providers to define external type constructors (\emph{tycons}) and their associated operators, implementing the logic governing their typechecking and translation to the (completely standard) IL using static metafunctions.
+For example, tycon structures can  express the type structure of tuples/records (and variants thereof), labeled sums, object systems, operations on format strings like \texttt{sprintf} and typechecked foreign function interfaces, all constructs that are, or would need to be, built in to other contemporary languages.
+A translation validation step  ensures type safety and hygiene, and perhaps most intriguingly, gives rise to a powerful modular reasoning principle: library providers can prove a broad class of metatheorems about the tycon structures they have defined in a ``closed world'' where no others need to be considered.
+All such metatheorems will necessarily continue to hold no matter which other tycon structures are also being used in a program, i.e.
+in the ``open world''.
+As in the module language, type abstraction plays a critical role in this process.%The module system that all of these constructs can be reasoned about separately and used in any combination.
+
+% There are three main mechanisms that we propose in this thesis.
+The first two have to do with concrete syntax, and are closely related, while the final one has to do with the type structure of the core language.
+% \begin{itemize}
+% \item \textbf{Typed Syntax Macros}.
+TODO: description + example (and a variant of them that further lowers syntactic cost in certain circumstances, \emph{type-specific languages}), introduced in Sec.
+\ref{sec:syntax}.
+% \item \textbf{Tycon Structures}, introduced in Sec. \ref{sec:extensible-semantics}. 
+% \end{itemize}
+% For each mechanism that we introduce, we will first demonstrate its expressive power by showing how constructs that are, or would need to be, built directly in to contemporary languages can in Verse be expressed in libraries.
+To demonstrate that these mechanisms are theoretically sound, we will then develop a formal specification and proofs of various important metatheoretic properties.
+The most interesting of these will be various \emph{modular reasoning principles} that guarantee that user-defined constructs can be reasoned about separately and then used together in any combination, without the possibility of syntactic or semantic conflict.
+
+%  %Only constructs for which the dynamic semantics can ultimately be defined by translation to the IL chosen can be expressed using the mechanisms we introduce (in practice, this would be exactly the same constraint necessary to guarantee that a compiler could implement a dialect specification).
+%Again, we will make these statements more precise as we continue.%The external language provides local type inference \cite{Pierce:2000:LTI:345099.345100}.
+% (for the purposes of this work).
+% %Such mechanisms are colloquially known as \emph{extension mechanisms} because the sorts of constructs they are used to express are those that would otherwise require extensions to the language itself.
+%Defining a new construct then corresponds to the usual notion of language extension.
+We will make this intuition more precise later.
+
+% % Rather than building derived syntax for certain privileged data structures, e.g.
+list syntax, as is nearly ubiquitious in dialects of ML, Verse allows users to express new derived syntax on their own, associating it with their own data structures.
+To guarantee that syntactic ambiguities cannot arise, Verse controls which delimiters can be used around derived syntax, but otherwise imposes no
+
+% %in languages without such mechanisms, expressing such constructs would require extensions to the language itself (a better qualifier might thus be \emph{extension-mitigating}).
+%We will make these notions more precise as we continue.
+%Thus, these mechanisms precisely capture the notions of language extension and language composition.
+
+% %The mechanisms we propose are designed to decrease the number of constructs that need to be built in \emph{a priori} by a language designer.
+However, there are still some basic syntactic and semantic decisions that must unavoidably be made.
+
+%  %Constructs that are built in to languages like ML, like $n$-ary tuples/records, labeled sums and others, can in Verse be realized as modes of use of this mechanism. 
+% %Languages that are organized around mechanisms like these are colloquially known as \emph{extensible languages}.
+
+% %Designing a modularly extensible language is a challenge because giving away too much control can easily make reasoning about program behavior difficult and lead to problems that arise only when extensions are used at the same time.
+For example, a library provider given free reign over syntax and semantics could introduce parsing conflicts, violate {type safety}, and  interfere with other extensions.
+We aim to design mechanisms that eliminate these possibilities, while providing a degree of expressive power sufficient to express a variety of constructs that are, or would need to be, built directly in to today's languages.
+%ll constructs that can be defined using a mechanism we will introduce can be reasoned about separately and composed arbitrarily.% With these mechanisms, we anticipate dialect formation being far less common.
+%Indeed, all of these problems have plagued previous  designs.
+
+% % Common dialects of ML that suffice for \cite{harper1997programming}. %We will .
+
+%    %Our challenge will be to design mechanisms that achieve these goals while also being powerful enough to express features that today require dialects.%Our challenge will be to design such extension mechanisms, while still retaining significant expressive power -- we wish to permit library-based expression of a range of features that today are, or would need to be, built into a system by its designers.%It can also make it more difficult to define new sorts of tools for working with the language because they cannot rely on there being a fixed abstract syntax (the canonical \emph{expression problem}).
+For programmers,
+% %It can make it difficult to understand and reason about the type of an unfamiliar term, a critical facet of program comprehension (the \emph{typing discipline problem}).
+
+% %So our goal will be to design extension mechanisms that handle these metatheoretic and composition-related issues while still  permitting library-based expression of a range of features that today are, or would need to be, built into a system by its designers.
+%More specifically, we will show how to delegate control, in a methodical manner, to \emph{static functions}, i.e.
+user-defined functions written in a \emph{static language}.
+By layering these mechanisms over a fixed typed internal language, constraining the static language appropriately and validating the code it is used to generate, we will retain key  metatheoretic properties and arrive at powerful modular reasoning principles.
+%By associating these static functions with type constructors, forming \emph{active type constructors},  we will further show how they can enjoy a usage profile essentially identical to that of built in features.
+% while maintaining key metatheoretic properties and modular reasoning principles.
+% By keeping the abstract syntax fixed and layering these mechanisms atop a fixed typed internal language (essentially lifting the first stages of compilation into the language), we avoid most aspects of the expression problem.
+%By delegating control to user-defined functions provided alongside user-defined type constructors (forming \emph{active type constructors}), we can give library providers substantially more control over these features of the system.
+% %We structure each mechanism that we introduce as a \emph{bidirectional type system} combined with an  \emph{elaboration semantics} targeting a fixed typed internal language. 
+% %By constraining these functions and enforcing critical abstraction barriers between extensions, we will ensure key metatheoretic properties of the language and system components and guarantee that extensions can be reasoned about modularly.
+%We call user-defined types that introduce new features into the system in this way \emph{active types}.
+% also highly expressive.
+% %But taking a \emph{language-internal approach} to implementing a feature is the most practical.
+If a feature can be realized by creatively using existing language constructs and distributed as a library, clients face fewer barriers to adoption because it is easy to integrate library-based features into existing projects gradually and granularly and they leverage well-understood and well-developed mechanisms.
+% % But taking this approach is often \emph{not} possible today %We call designs \emph{monolithic programming systems}.
+
+% %To realize a new abstraction or system behavior, such experts can consider either a \emph{language-internal approach}, where they work within an existing language and distribute their solutions as libraries, or a \emph{language-external approach}, where they create a new, distinct programming system (often centered around what has come to be called a new \emph{domain-specific language} \cite{dsl}) or extend an existing system by some mechanism that is not part of the language itself, such as an extension mechanism supported by a {particular} compiler, editor or other tool.
+
+% %The result of this work will be the design of a \emph{modularly extensible programming language} called Verse.
+and end with a brief foray into editor services in Sec.
+\ref{sec:editor-services}.\todo{fix sections}
+
+% \section{Extensible Semantics}\label{sec:extensible-semantics}
+% As a minimal example, consider System \textbf{F}, a.k.a.
+the polymorphic lambda calculus (perhaps the simplest general-purpose language) \cite{girard1971extension,Reynolds94anintroduction,pfpl}.
+It specifies type variables and two type constructors, written abstractly $\mathtt{arr}$ and $\mathtt{all}$:
+
+% \[
+% \begin{array}{llllll}
+% \textbf{Sort} & & & \textbf{Abstract Form} & \textbf{Typeset Form} & \textbf{Description}\\
+% \mathsf{Type} & \tau & ::= & t & t & \text{type variable}\\
+% & & & \mathtt{arr}(\tau; \tau) & \tau \rightarrow \tau & \text{function type}\\
+% & & & \mathtt{all}(t.\tau) & \forall t.\tau & \text{type quantification}
+% \end{array}
+% \]
+% and value variables and four term constructors:
+% \[
+% \begin{array}{llllll}
+% \textbf{Sort} & & & \textbf{Abstract Form} & \textbf{Typeset Form} & \textbf{Description}\\
+% \mathsf{Exp} & \iota & ::= & x & x & \text{value variable}\\
+% & & & \mathtt{lam}[\tau](x.\iota) & \lambda x{:}\tau.\iota & \text{value abstraction}\\
+% & & & \mathtt{ap}(\iota; \iota) & \iota(\iota) & \text{application}\\
+% & & & \mathtt{Lam}(t.\iota) & \Lambda t.\iota & \text{type abstraction}\\
+% & & & \mathtt{Ap}[\tau](\iota) & \iota\mathtt{[}\tau\mathtt{]} & \text{type application}
+% \end{array}
+% \]
+
+% If we wanted to express record types in \textbf{F}, derived syntax would not be enough (the static semantics of \textbf{F} provide no way to express row labels, amongst other issues).
+We could, however, develop a dialect that built in the type and term operators relevant to records, specifying its static and dynamic semantics as a type-directed translation to the polymorphic lambda calculus (using the standard Church-style encoding for records).
+More formally, assuming that metavariables $e$, $\sigma$ and $\Upsilon$ range over terms, types and typing contexts of our dialect, and $\iota$, $\tau$ and $\Gamma$  range over the terms, types and typing contexts of \textbf{F}, the main judgements in the specification of our dialect would take the following form:
+% \[\begin{array}{ll}
+% \textbf{Judgement} & \textbf{Pronunciation}\\
+% \Upsilon \vdash e : \sigma \leadsto \iota & \text{Under typing context $\Upsilon$, term $e$ has type $\sigma$ and translation $\iota$.}\\
+% \Upsilon \vdash \sigma \leadsto \tau & \text{Under typing context $\Upsilon$, type $\sigma$ is valid and has translation $\tau$.}\\
+% \vdash \Upsilon \leadsto \Gamma & \text{Typing context $\Upsilon$ is valid and has translation $\Gamma$.}
+% \end{array}\]
+% \todo{couple more sentences} %Constructs specified by a type-directed translation like this can be considered \emph{derived semantic constructs}.
+
+% \subsubsection{Example 2: Regular Strings}\label{sec:rstr}
+% The example above deals with how regular expression patterns are  encoded, introduced and reasoned about.
 Going further, programmers may also benefit from a semantics where they can statically constrain strings to be within a particular regular language \cite{sanitation-psp14}.
 For example, a programmer might want to ensure that the arguments to a function that creates a database connection given a username and password are alphanumeric strings, ideally including this specification directly in the type of the  function:
-\begin{lstlisting}[numbers=none]
-type alphanumeric = rstring /SURL[A-Za-z0-9]+EURL/
-val connect : (alphanumeric * alphanumeric) -> DBConnection
-\end{lstlisting}
-
-This example requires that the language include a {type constructor}, \verb|rstring|, indexed by a statically known regular expression, written using the syntax described above.
+% \begin{lstlisting}[numbers=none]
+% type alphanumeric = rstring /SURL[A-Za-z0-9]+EURL/
+% val connect : (alphanumeric * alphanumeric) -> DBConnection
+% \end{lstlisting}
+
+% This example requires that the language include a {type constructor}, \verb|rstring|, indexed by a statically known regular expression, written using the syntax described above.
 Ideally, we would be able to use standard string literal syntax for such regular strings as well, e.g.
-\begin{lstlisting}[numbers=none]
-let connection = connect("admin", "password")
-\end{lstlisting}
-
-To be useful, the language would also need a static semantics for standard operations on regular strings.
+% \begin{lstlisting}[numbers=none]
+% let connection = connect("admin", "password")
+% \end{lstlisting}
+
+% To be useful, the language would also need a static semantics for standard operations on regular strings.
 For example, concatenating two alphanumeric strings  should result in an  alphanumeric string.
@@
 
-Regular strings might go beyond simply \emph{refining} standard string types (i.e.
+% Regular strings might go beyond simply \emph{refining} standard string types (i.e.
 specifying verification conditions atop an existing semantics \cite{Freeman91}).
 For example, they might define an operation like \emph{captured group projection} that has no analog over standard strings:
-\begin{lstlisting}[numbers=none]
-let example : rstring /SURL(\d\d\d)-(\d\d\d\d)EURL/ = "555-5555"
-let group0 (* : rstring /\d\d\d/ *) = example#0
-\end{lstlisting}
-It should be possible to give this operation a cost of $\mathcal{O}(1)$.
+% \begin{lstlisting}[numbers=none]
+% let example : rstring /SURL(\d\d\d)-(\d\d\d\d)EURL/ = "555-5555"
+% let group0 (* : rstring /\d\d\d/ *) = example#0
+% \end{lstlisting}
+% It should be possible to give this operation a cost of $\mathcal{O}(1)$.
 %From a language implementation perspective, this requires that different regular string types  have different representations, also not possible when using a refinement system over strings.
@@
 
-\subsubsection{Example 3: Labeled Products with Functional Update Operators}\label{sec:lprod}
-The simplest way to specify the semantics of product types is to specify only nullary and binary products \cite{pfpl}.
+% \subsubsection{Example 3: Labeled Products with Functional Update Operators}\label{sec:lprod}
+% The simplest way to specify the semantics of product types is to specify only nullary and binary products \cite{pfpl}.
 However, in practice, many more general but also more complex variations on product types are built in to various dialects of ML and other languages, e.g.
 $n$-ary tuples, labeled tuples,
-records (identified up to reordering), 
-records with width and depth coercions \cite{Cardelli:1984:SMI:1096.1098} and records with functional update operators (also called \emph{extensible records}) \cite{ocaml-manual}.
+% records (identified up to reordering), 
+% records with width and depth coercions \cite{Cardelli:1984:SMI:1096.1098} and records with functional update operators (also called \emph{extensible records}) \cite{ocaml-manual}.
 {The Haskell wiki notes that ''extensible records  are not implemented in GHC.
@@
 [...] As a result, nothing much happens.'' \cite{GHCFAQ}}%,
- %mutable fields \cite{ocaml-manual}, 
-%field delegation \cite{atlang-gpce14} \todo{gpce submission} 
-%and 
-% ``methods'' (i.e. pure objects) \cite{TSLs}
-
-We would ideally like to avoid needing the language designer to decide \emph{a priori} on just one privileged point in this large design space.
+%  %mutable fields \cite{ocaml-manual}, 
+% %field delegation \cite{atlang-gpce14} \todo{gpce submission} 
+% %and 
+% % ``methods'' (i.e. pure objects) \cite{TSLs}
+
+% We would ideally like to avoid needing the language designer to decide \emph{a priori} on just one privileged point in this large design space.
 Instead, the language designer might define only nullary and binary products, and include an extension mechanism that makes it possible for these other variations on products to be defined as libraries and  used together without conflict.
@@
 An example of a labeled product type (constructed by the type constructor \lstinline{lprod}) classifying conference papers might be:
-\begin{lstlisting}[numbers=none]
-type Paper = lprod {
-  title : rstring /SURL.+EURL/,
-   conf : rstring /SURL[A-Z]+ \d\d\d\dEURL/
-}
-\end{lstlisting}
-
-The row ordering should make it possible to introduce values of this type either with or without explicit labels, e.g.
-\begin{lstlisting}[numbers=none]
-fun make_paper(t : rstring /SURL.+EURL/) : Paper = {title=t, conf="EXMPL 2015"}
-\end{lstlisting}
-should be equivalent to
-\begin{lstlisting}[numbers=none]
-fun make_paper(t : rstring /SURL.+EURL/) : Paper = (t, "EXMPL 2015")
-\end{lstlisting}
-(note that we aim to be able to unambiguously re-use standard record and tuple syntax.)
-
-We should then be able to project out rows by providing a positional index or a label:
-\begin{lstlisting}[numbers=none]
-let test_paper = make_paper "Test Paper"
-test_paper#0
-test_paper#conf
-\end{lstlisting}
-
-We might also want our labeled tuples to support functional update operators. For example, an operator that dropped a row might be used like this:
-\begin{lstlisting}[numbers=none]
-let title_only (* : lprod {title : rstring /.+/} *) = test_paper.drop[conf]
-\end{lstlisting}
-An operation that added a row, or updated an existing row, might be used like this:
-\begin{lstlisting}[numbers=none]
-fun with_author(p : Paper, a : rstring /SURL.+EURL/) = p.ext(author=a)
-\end{lstlisting}
-% \subsection{Language-External Approaches}\label{external-approaches}
-
-% In situations like these where a library-based approach is not satisfactory, providers must today take a \emph{language-external approach}, specifying a new system dialect, reasoning about its metatheory monolithically and implementing it using a {compiler generator} \cite{brooker1963compiler}, {language workbench} \cite{erdweg2013state}, {DSL framework} \cite{fowler2010domain}, an extension mechanism for a {particular} compiler\footnote{Compilers that modify, or allow modification of, the semantics of their base language, rather than simply permitting semantics-preserving optimizations, should be considered a pernicious means for creating dialects.
-% %That is, some programs that purport to be written in C, Haskell or Standard ML are actually written in compiler-specific dialects of these languages.
-% }, editor or other tool, or by forking an existing system implementation.
-
-% For example, a researcher interested in providing the regular expression related features just described (let us refer to these collectively as \texttt{R}) might design a new system with built-in support for them, perhaps basing it on an existing system containing standard general-purpose features, \texttt{G} (e.g.
+% \begin{lstlisting}[numbers=none]
+% type Paper = lprod {
+%   title : rstring /SURL.+EURL/,
+%    conf : rstring /SURL[A-Z]+ \d\d\d\dEURL/
+% }
+% \end{lstlisting}
+
+% The row ordering should make it possible to introduce values of this type either with or without explicit labels, e.g.
+% \begin{lstlisting}[numbers=none]
+% fun make_paper(t : rstring /SURL.+EURL/) : Paper = {title=t, conf="EXMPL 2015"}
+% \end{lstlisting}
+% should be equivalent to
+% \begin{lstlisting}[numbers=none]
+% fun make_paper(t : rstring /SURL.+EURL/) : Paper = (t, "EXMPL 2015")
+% \end{lstlisting}
+% (note that we aim to be able to unambiguously re-use standard record and tuple syntax.)
+
+% We should then be able to project out rows by providing a positional index or a label:
+% \begin{lstlisting}[numbers=none]
+% let test_paper = make_paper "Test Paper"
+% test_paper#0
+% test_paper#conf
+% \end{lstlisting}
+
+% We might also want our labeled tuples to support functional update operators. For example, an operator that dropped a row might be used like this:
+% \begin{lstlisting}[numbers=none]
+% let title_only (* : lprod {title : rstring /.+/} *) = test_paper.drop[conf]
+% \end{lstlisting}
+% An operation that added a row, or updated an existing row, might be used like this:
+% \begin{lstlisting}[numbers=none]
+% fun with_author(p : Paper, a : rstring /SURL.+EURL/) = p.ext(author=a)
+% \end{lstlisting}
+% % \subsection{Language-External Approaches}\label{external-approaches}
+
+% An important point is that labeled products, as well as other ``record-like'' constructs, can  be expressed by a \emph{type-directed translation} targeting our IL (and indeed, many compilers take advantage of facts like this to simplify the number of cases that must be considered downstream).
+%it is inaccurate to state that they are merely syntactic sugar over such a language.
+%Object systems are perhaps even more diverse across languages\todo{citation}.
+% %Other dialects are justified by a desire to provide more special-purpose semantic constructs. For example, 
+% Substantially more sophisticated examples require at most only a very slightly more capable intermediate language (the IL of popular virtual machines, even those with flaws like ubiquitous \texttt{null} references, e.g.
+JVM or CIL bytecode, are also sufficient if used strictly as translation targets).
+%Once again, these languages could all be specified by type-directed translation to a common intermediate language with comparatively few constructs , but generally not by a purely syntactic desugaring.
+
+% % In situations like these where a library-based approach is not satisfactory, providers must today take a \emph{language-external approach}, specifying a new system dialect, reasoning about its metatheory monolithically and implementing it using a {compiler generator} \cite{brooker1963compiler}, {language workbench} \cite{erdweg2013state}, {DSL framework} \cite{fowler2010domain}, an extension mechanism for a {particular} compiler\footnote{Compilers that modify, or allow modification of, the semantics of their base language, rather than simply permitting semantics-preserving optimizations, should be considered a pernicious means for creating dialects.
+% % %That is, some programs that purport to be written in C, Haskell or Standard ML are actually written in compiler-specific dialects of these languages.
+% % }, editor or other tool, or by forking an existing system implementation.
+
+% % For example, a researcher interested in providing the regular expression related features just described (let us refer to these collectively as \texttt{R}) might design a new system with built-in support for them, perhaps basing it on an existing system containing standard general-purpose features, \texttt{G} (e.g.
 the features of Standard ML).
@@
 
-% Although these systems are valid ``proofs of concept'', this is a rather weak notion because no abstraction will be used entirely in isolation.
+% % Although these systems are valid ``proofs of concept'', this is a rather weak notion because no abstraction will be used entirely in isolation.
 Such an approach gives us no rigorously justifiable  reason to believe that it can safely and naturally be merged with others specified or implemented in the same way.
@@
 % and there is no formal, general mechanism for merging these systems.
-% We argue that, due to these problems of composition, taking a language-external approach to realizing a new feature should be considered harmful and avoided whenever possible.
-
-% \begin{figure}
-% \begin{center}
-% \includegraphics[scale=.48]{approaches.pdf}
-% \end{center}
-% \vspace{-20px}
-% \caption{\small (a) When taking a language-external approach, new features are packaged together into separate languages and tools.
+% % We argue that, due to these problems of composition, taking a language-external approach to realizing a new feature should be considered harmful and avoided whenever possible.
+
+% % \begin{figure}
+% % \begin{center}
+% % \includegraphics[scale=.48]{approaches.pdf}
+% % \end{center}
+% % \vspace{-20px}
+% % \caption{\small (a) When taking a language-external approach, new features are packaged together into separate languages and tools.
 (b) We propose an approach where there is one extensible host language and the compile-time and edit-time logic governing new constructs is expressed within safely composable libraries.}
-% \label{approaches}
-% %\vspace{-10px}
-% \end{figure}
-%  %
-% % More specifically, there is no guarantee of \textbf{orthogonality} or even \textbf{interoperability}.% This has limited the broad adoption of these kinds of innovations.%This latter method couples the semantics of the feature to the implementation details of a particular tool.
+% % \label{approaches}
+% % %\vspace{-10px}
+% % \end{figure}
+% %  %
+% % % More specifically, there is no guarantee of \textbf{orthogonality} or even \textbf{interoperability}.% This has limited the broad adoption of these kinds of innovations.%This latter method couples the semantics of the feature to the implementation details of a particular tool.
 Because the use of one implementation entails a different semantics for the feature than another, the extended tool acts, \emph{de facto}, as a distinct system for our purposes.
 
-% %\paragraph{Orthogonality} Features implemented by language-external means like these cannot be adopted individually, but instead are only available coupled to a fixed collection of other features.
+% % %\paragraph{Orthogonality} Features implemented by language-external means like these cannot be adopted individually, but instead are only available coupled to a fixed collection of other features.
 This makes adoption more costly when these incidental features are  not desirable or are insufficiently developed (``toy languages''), or when the features bundled with a different language or tool are simultaneously desirable.
 
-% % Recent evidence indicates that this is one of the major barriers preventing research  from being driven into practice.
+% % % Recent evidence indicates that this is one of the major barriers preventing research  from being driven into practice.
 For example, developers prefer high-level language-integrated parallel programming abstractions that provide  stronger semantic guarantees  \cite{cave2010comparing}, but library-based approximations are far more widely adopted because the ``parallel programming languages'' privilege only a few  specialized abstractions at the language level.
@@
 
-% %\item Developing a new language and its associated tools places a significant development burden on providers who may wish only to promote a few core innovations, although tools like compiler generators, language workbenches and easy-to-extend tools can decrease this burden.
-% %\item
-
-% %Clients seem to prioritize the ability to choose different features for different portions of an application. 
-% %If calling between languages were safe and easy, then using a variety of specialized languages and associated tools might be less problematic. In fact, s
-% %Recognizing the limitations of relying on monolithic collection of primitives, some researchers have advocated instead for a model where multiple languages used within a single application, calling it the \emph{language-oriented approach} to software development \cite{languageoriented}.
-
-% % \paragraph{Interoperability} Even in cases where, for each component of a software system, a programming system considered entirely satisfactory by its developers is available (e.g.
+% % %\item Developing a new language and its associated tools places a significant development burden on providers who may wish only to promote a few core innovations, although tools like compiler generators, language workbenches and easy-to-extend tools can decrease this burden.
+% % %\item
+
+% % %Clients seem to prioritize the ability to choose different features for different portions of an application. 
+% % %If calling between languages were safe and easy, then using a variety of specialized languages and associated tools might be less problematic. In fact, s
+% % %Recognizing the limitations of relying on monolithic collection of primitives, some researchers have advocated instead for a model where multiple languages used within a single application, calling it the \emph{language-oriented approach} to software development \cite{languageoriented}.
+
+% % % \paragraph{Interoperability} Even in cases where, for each component of a software system, a programming system considered entirely satisfactory by its developers is available (e.g.
 a team goes through the trouble of implementing \verb|G+R+P| by reading papers about \verb|G+R| and \verb|G+P| and disentangling orthogonality-related issues), there remains a problem at any interface between  components written using a different combination of features.
@@
 
-% % One strategy taken by proponents of a {language-oriented approach} \cite{journals/stp/Ward94} to partially address the interoperability problem is to  target an established intermediate language and use its constructs as a common language for communication between components written in different languages.
+% % % One strategy taken by proponents of a {language-oriented approach} \cite{journals/stp/Ward94} to partially address the interoperability problem is to  target an established intermediate language and use its constructs as a common language for communication between components written in different languages.
 Scala \cite{200464/IC} and F\# \cite{pickering2007foundations} are examples of prominent general-purpose languages that have taken this approach, and most DSL frameworks also rely on this strategy.
@@
 
-% % %This approach only works well when new languages consist of constructs that can also be expressed safely and almost as naturally in the common language.
-% % %But many of the most innovative constructs found in modern languages (often, those that justify their creation) are difficult to define in terms of existing constructs in ways that guarantee all necessary invariants are statically maintained and that do not require large amounts boilerplate code and run-time overhead.
-% % As a simple example with significant contemporary implications, F\#'s type system does not admit \verb|null| as a value for any type both defined and used within F\# code, but maintaining this sensible internal invariant still requires dynamic  checks because the stricter typing rules of F\# do not apply when F\# data structures are constructed by other languages on the Common Language Infrastructure (CLI) like C\# or SML.NET.
+% % % %This approach only works well when new languages consist of constructs that can also be expressed safely and almost as naturally in the common language.
+% % % %But many of the most innovative constructs found in modern languages (often, those that justify their creation) are difficult to define in terms of existing constructs in ways that guarantee all necessary invariants are statically maintained and that do not require large amounts boilerplate code and run-time overhead.
+% % % As a simple example with significant contemporary implications, F\#'s type system does not admit \verb|null| as a value for any type both defined and used within F\# code, but maintaining this sensible internal invariant still requires dynamic  checks because the stricter typing rules of F\# do not apply when F\# data structures are constructed by other languages on the Common Language Infrastructure (CLI) like C\# or SML.NET.
 This is not an issue exclusive to intermediate languages that make regrettable choices regarding \verb|null|, however.
@@
 Moreover, it also cannot naturally consume F\# data structures, despite having a rather similar syntax and semantics in most ways (both languages directly descend from ML).
-% %In Scala, traits that have default method implementations are difficult to implement from Java or other JVM languages and the workaround can break if the trait is modified \cite{scalatraitinterop}.
-% %In some cases, desirable features must be omitted entirely due to concerns about interoperability.
+% % %In Scala, traits that have default method implementations are difficult to implement from Java or other JVM languages and the workaround can break if the trait is modified \cite{scalatraitinterop}.
+% % %In some cases, desirable features must be omitted entirely due to concerns about interoperability.
 F\#, for example, aimed to retain source compatibility with Ocaml code, but due to the need for bidirectional interoperability with CLI languages, it does not support features like polymorphic variants, modules or functors \cite{ocaml-manual} because they have no apparent analogs in the type system of the CLI.
-% %\end{itemize}
-
-% %\subsection{Language-Integrated Approaches}\label{language-integrated-approaches}
-
-% %Such libraries have been called \emph{active libraries}  \cite{activelibraries}.
+% % %\end{itemize}
+
+% % %\subsection{Language-Integrated Approaches}\label{language-integrated-approaches}
+
+% % %Such libraries have been called \emph{active libraries}  \cite{activelibraries}.
 %Features implemented within active libraries can be imported individually, unlike features implemented by external means, giving us a potential means to avoid the problems of orthogonality and interoperability just described.
 
-% % We must proceed with caution, however: critical issues having to do with {safety} must be overcome before language-integrated extension mechanisms can be introduced into a system.
+% % % We must proceed with caution, however: critical issues having to do with {safety} must be overcome before language-integrated extension mechanisms can be introduced into a system.
 If too much control over  these core features of the system is given  to developers, the system's metatheory may become weak.
-% % %For example, an extension could weaken important metatheoretic guarantees previously provided by the system. 
-% % For example, parsing ambiguities might arise if the syntax can be modified arbitrarily.
+% % % %For example, an extension could weaken important metatheoretic guarantees previously provided by the system. 
+% % % For example, parsing ambiguities might arise if the syntax can be modified arbitrarily.
 Type safety  may not hold if the static and dynamic semantics of the language can be modified or extended arbitrarily from within libraries.
@@
 %Resolving these kinds of ambiguities requires significantly more expertise with parser technology than using the syntax itself does.
-% % These issues have plagued previous attempts to design language-integrated extensibility mechanisms.\todo{I can include a more detailed related work section if requested.}% We will briefly review some of these attempts below, then return to our approach.% To prevent them, our mechanisms will organize  extension logic around types to guarantee that extensions are both safe in isolation and also safely composable in any combination.
-
-%  %This represents a minimalist approach to system design -- the conventional distinction between built-in and user-defined constructs is blurred and most features of the system are orthogonally implemented as {libraries}, rather than by the maintainers of the system.
-
-% %The mechanisms we describe will do so primarily by delimiting the scope of an extension to expressions of a single user-defined type or family of types.
-
-% %This can be thought of as a more pernicious form of the conflict that arises when two globally-accessible constructs are given the same name.
+% % % These issues have plagued previous attempts to design language-integrated extensibility mechanisms.\todo{I can include a more detailed related work section if requested.}% We will briefly review some of these attempts below, then return to our approach.% To prevent them, our mechanisms will organize  extension logic around types to guarantee that extensions are both safe in isolation and also safely composable in any combination.
+
+% %  %This represents a minimalist approach to system design -- the conventional distinction between built-in and user-defined constructs is blurred and most features of the system are orthogonally implemented as {libraries}, rather than by the maintainers of the system.
+
+% % %The mechanisms we describe will do so primarily by delimiting the scope of an extension to expressions of a single user-defined type or family of types.
+
+% % %This can be thought of as a more pernicious form of the conflict that arises when two globally-accessible constructs are given the same name.
 n languages without universal namespacing mechanisms (e.g.
@@
 
-% %The extension mechanism\todo{elaborate on safety requirements + tension between expressiveness and safety, merge with next paragraph}.
+% % %The extension mechanism\todo{elaborate on safety requirements + tension between expressiveness and safety, merge with next paragraph}.
 must be expressive enough to allow users to associate rich run-time, compile-time and edit-time behaviors with user constructs directly, while being sufficiently restrictive to maintain the global safety properties of the language and system as a whole, and to ensure that constructs cannot interfere with one another.

### 2015-03-26 thesis-proposal/problem.tex
updated proposal

@@
 
-Well-designed general-purpose programming languages give library providers the power to  modularly express a wide variety of useful constructs using a comparatively small set of primitives.
+%Well-designed general-purpose programming languages give library providers the power to  modularly express a wide variety of useful constructs using a comparatively small set of primitives.
 %New primitives, and thus new dialects of these languages, should arise very rarely.
@@
 Instead, its users are able to modularly  express a wide variety of useful constructs in libraries, using the comparatively small set of primitives that the language builds in.
-There has been substantial progress on many of the foundational semantic issues that are relevant to the design of general-purpose languages over the past  decades, much of it driven by advances in type theory.
-However, when more pragmatic language design issues are brought into consideration, a broadly agreeable set of primitives remains elusive, as evidenced by the diverse array of dialects that surround around all major con\-temporary languages.
+Over the past several decades, researchers have developed a substantial understanding of  the theoretical issues relevant to the design of programming languages, in large part by constructing and studying the metatheoretic properties of minimal typed lambda calculi.
+These minimal calculi are clearly not directly suitable for real-world software development, so language designers must expose a syntax and type structure of a more elaborate design to programmers.
+%syntactic and semantic primitives that capture common idioms naturally and precisely.
+Language designers have many viable choices at this level, as evidenced by the diverse array of dialects that continue to proliferate around all major languages.
 For example, let us consider the many dialects of ML.
-Though they generally agree on the importance of a powerful module system atop a core language based conceptually on the   polymorphic lambda calculus with sums, products and recursive types, they all introduce various syntactic and semantic extensions and tweaks of their own novel design (we discuss a number of examples below).
-Tools for  constructing new ``domain-specific'' language dialects also continue to proliferate.
-% The situation is similar across all major language lineages.
-Why are well-informed programmers and researchers so often tempted to construct  dialects of a language like ML? 
+Though they generally agree on the importance of a powerful module system atop a core language based conceptually on the   polymorphic lambda calculus with recursive sum and product types, they all introduce various primitive syntactic and semantic extensions and tweaks of their own novel design (we cover several examples below).
+%Tools for  constructing new ``domain-specific'' dialects also continue to proliferate. % The situation is similar across all major language lineages. 
+Tools for  constructing ``domain-specific'' dialects contribute further to this proliferation of language primitives.
+This is alarming: why is it that well-informed programmers and researchers are so often unable to express the constructs that they need in libraries, as modes of use of a small number of ``general-purpose'' language primitives?
 %Let us investigate why this is.
@@
 For example,  we will consider regular expression patterns expressed using abstract data types (in ML, a mode of use of the module system) in Section \ref{sec:syntax}.
-This gives us precisely the semantics that we need, but the syntactic cost  is not quite ideal.
+This will give  us precisely the semantics that we need, but the syntactic cost will not be ideal.
 In situations like this, library providers are often tempted to construct a \emph{derived syntactic dialect}, i.e.
@@
 For example, Ur/Web is a derived syntactic dialect of Ur (itself descended from ML) that builds in derived syntax for SQL queries, HTML elements and other datatypes common in web programming \cite{conf/popl/Chlipala15}.
-Indeed, nearly all major languages primitively build in derived syntax for constructs that are otherwise expressed in the standard library, e.g.
+In fact, nearly all major languages primitively build in derived syntax for constructs that are otherwise expressed in the standard library, e.g.
 derived list syntax in Standard ML (SML) \cite{harper1997programming,mthm97-for-dart}.
@@
 
-The problem with this practice of using language dialects as vehicles for new syntactic and semantic primitives is that it is, in an important sense,  anti-modular: a program written in one dialect cannot, in general, safely and idiomatically interface with libraries written in a different dialect.
+The reason why this practice of using dialects as vehicles for new syntactic and semantic primitives is alarming is that it is, in an important sense,  anti-modular: a program written in one dialect cannot, in general, safely and idiomatically interface with libraries written in a different dialect.
 %At best, one can hope that the compilers for the decidedlytwo languages target a common intermediate language,
@@
 
-Dialects do, however, have an indirect impact on the practice of programming because the language designers that control comparatively popular languages, like Ocaml and Scala, can occasionally be convinced to incorporate  ideas from dialects  into backwards compatible language revisions.
+Dialects can, however, have an indirect impact on large-scale software development because the language designers that control comparatively popular languages, like Ocaml and Scala, can occasionally be convinced to incorporate  ideas from dialects  into backwards compatible language revisions.
 %These decisions are increasingly influenced by community processes, e.g.
@@
 Moreover, because there appears to be a ``long tail'' of potentially useful primitives (for derived syntactic primitives, our group has gathered initial data speaking to this \cite{TSLs}),  primitives that are only situationally useful, or that trigger aesthetic disagreements between different factions of the community, are still often left languishing in ``toy'' dialects.
-A conservative approach to incorporating new primitives is only sensible because once a primitive is introduced, it becomes entrenched as-is and monopolizes ``syntactic resources'', as we will discuss below.
-Recalling the words of  Reynolds, which are clearly as relevant today as they were nearly half a century ago \cite{Reynolds70}:%Because there is no data about how useful a construct is in practice until it is included in a language like this, decisions about which constructs to include are often informed by little more than intuition.
+A conservative approach to incorporating new primitives is sensible because once a primitive is introduced, it becomes entrenched as-is and monopolizes ``syntactic resources'', as we will discuss below.Recalling the words of  Reynolds, which are clearly as relevant today as they were nearly half a century ago \cite{Reynolds70}:%Because there is no data about how useful a construct is in practice until it is included in a language like this, decisions about which constructs to include are often informed by little more than intuition.
 %This approach is antithetical to the ideal of a truly \emph{general-purpose language} described at the beginning of this section.
@@
 
-\newpage
- There are two possible paths forward for the subset of the language design community interested in keeping general-purpose languages small and free of \emph{ad hoc} primitives.
+\newpage There are two possible paths forward for the subset of the language design community interested in keeping general-purpose languages small and free of \emph{ad hoc} primitives.
 One, exemplified (arguably) by Standard ML, is to generally eschew the introduction of new primitives and settle on a set of primitives that sit at a ``sweet spot'' in the overall language design space, accepting that in some circumstances, these primitives trade away expressive power or have high syntactic cost.

### 2015-04-08 thesis-proposal/problem.tex
proposal

@@
 Instead, its users are able to modularly  express a wide variety of useful constructs in libraries, using the comparatively small set of primitives that the language builds in.
-Over the past several decades, researchers have developed a substantial understanding of  the theoretical issues relevant to the design of programming languages, in large part by constructing and studying the metatheoretic properties of minimal typed lambda calculi.
-These minimal calculi are clearly not directly suitable for real-world software development, so language designers must expose a syntax and type structure of a more elaborate design to programmers.
+Functional programming languages like Standard ML (SML) and Haskell descend directly from elegant typed lambda calculi, diverging mainly in that they strive to present a more pragmatic concrete syntax and type structure to programmers.
+In other words, their design considers important human factors in addition to fundamental metatheoretic issues.
+For example, both languages build in record types, generalizing the nullary and binary product types more common in simpler calculi, because labels are cognitively useful.
+%Minimal typed lambda calculi play a central role in the study of programming languages, and serve as direct inspiration for  functional programming languages like ML.
+to be useful in practice, it must present a decidedly more elaborate concrete syntax and type structure to programmers, designed considering various human factors in addition to fundamental metatheoretic issues.
+For example, while it is quite common to study minimal calculi that build in only nullary and binary product types, Standard ML (SML) builds in record types, because labels are cognitively useful.
 %syntactic and semantic primitives that capture common idioms naturally and precisely.
-Language designers have many viable choices at this level, as evidenced by the diverse array of dialects that continue to proliferate around all major languages.
-For example, let us consider the many dialects of ML.
-Though they generally agree on the importance of a powerful module system atop a core language based conceptually on the   polymorphic lambda calculus with recursive sum and product types, they all introduce various primitive syntactic and semantic extensions and tweaks of their own novel design (we cover several examples below).
+
+Ideally, transitioning from a minimal calculus to a pragmatic programming language would require introducing only a limited number of primitive ``embellishments'' like this.
+The community around the language would then be able to redirect its efforts exclusively toward the development of a wide variety of useful libraries.
+Alas, a language design that achieves this ideal has proven elusive, as evidenced by the diverse array of {dialects} that continue to proliferate around all major languages, functional languages included.
+% For example, let us consider the many dialects of ML.
+Though they generally agree on the importance of a powerful module system atop a core language based conceptually on the   polymorphic lambda calculus with recursive sum and product types, they all introduce various primitive syntactic and semantic extensions and tweaks of their own design (we cover several examples below).
 %Tools for  constructing new ``domain-specific'' dialects also continue to proliferate. % The situation is similar across all major language lineages. 
-Tools for  constructing ``domain-specific'' dialects contribute further to this proliferation of language primitives.
-This is alarming: why is it that well-informed programmers and researchers are so often unable to express the constructs that they need in libraries, as modes of use of a small number of ``general-purpose'' language primitives?
+In fact, tools for constructing new ``domain-specific'' dialects seem only to be increasingly popular. 
+{This calls for an investigation}: why is it that well-informed programmers and researchers are so often unsatisfied with expressing the constructs that they need in libraries, as modes of use of the ``general-purpose'' functional primitives available today?
 %Let us investigate why this is.
@@
 
-Perhaps the most common  reason is simply that, for human {programmers}, the \emph{syntactic cost} of expressing a construct of interest using general-purpose primitives is not always ideal.
+Perhaps the most common  reason may simply be that the \emph{syntactic cost} of expressing a construct of interest using contemporary general-purpose primitives is not always ideal.
 %Syntactic cost is often assessed qualitatively \cite{green1996usability}, though quantitative metrics can be defined.
 For example,  we will consider regular expression patterns expressed using abstract data types (in ML, a mode of use of the module system) in Section \ref{sec:syntax}.
-This will give  us precisely the semantics that we need, but the syntactic cost will not be ideal.
-In situations like this, library providers are often tempted to construct a \emph{derived syntactic dialect}, i.e.
+This  will give  us precisely the semantics that we need, but syntactically, it will leave something to be desired.
+To avoid this syntactic cost, library providers are frequently tempted to construct a \emph{derived syntactic dialect}, i.e.
 a dialect specified by context-free elaboration to the existing language, that introduces derived syntax specific to their library.
 For example, Ur/Web is a  derived syntactic dialect of Ur (itself descended from ML) that builds in derived syntax for SQL queries, HTML elements and other datatypes common in web programming \cite{conf/popl/Chlipala15}.
-In fact, nearly all major languages primitively build in derived syntax for constructs that are otherwise expressed in the standard library, e.g.
-derived list syntax in Standard ML (SML) \cite{harper1997programming,mthm97-for-dart}.
-Tools like Camlp4 \cite{ocaml-manual} and Sugar* \cite{erdweg2011sugarj,erdweg2013framework}, which we will return to later, help lower the engineering costs of constructing derived syntactic dialects, contributing to their proliferation.% contributing to their proliferation.
+In fact, nearly all major programming languages primitively build in derived syntax for constructs that can otherwise be expressed in the standard library, e.g.
+derived list syntax in SML \cite{harper1997programming,mthm97-for-dart}.
+Tools like Camlp4 \cite{ocaml-manual} and Sugar* \cite{erdweg2011sugarj,erdweg2013framework} (which we will return to later) help lower the engineering costs of constructing such dialects, contributing to their proliferation.% contributing to their proliferation.
 
@@
 
-More advanced dialects introduce  new type structure, going beyond what is possible by a context-free elaboration to an existing language.
-For example, SML goes beyond the simple nullary and binary product types common in simpler calculi, instead exposing record types.
+More advanced dialects introduce  new type structure, going beyond what is possible by context-free elaboration to the existing language.
+For example, as  mentioned above, Standard ML goes beyond the nullary and binary product types found in simpler calculi, instead exposing record types.
 Other dialects have explored  ``record-like'' primitives that support functional update and extension operators, width and depth coercions (sometimes implicit)%\cite{Cardelli:1984:SMI:1096.1098}
 , mutable rows, methods, prototypic dispatch and various other embellishments.
-Ocaml builds in a more  sophisticated object system, polymorphic variants and  conveniences like logic for typechecking operations that use format strings, e.g.
+Ocaml builds in a  sophisticated object system, polymorphic variants and  conveniences like logic for typechecking operations that use format strings, e.g.
 $\mathtt{sprintf}$ \cite{ocaml-manual}.
@@
 Alice ML builds in futures and promises \cite{AliceLookingGlass}.
-MLj builds in essentially the entire Java programming language, to support typesafe interoperation with existing Java code \cite{Benton:1999:IWW:317636.317791}.
+MLj builds in essentially the entire Java programming language, to support typesafe interoperation with existing Java programs \cite{Benton:1999:IWW:317636.317791}.
 Indeed, perusal of the proceedings of any major programming languages conference will typically yield several more examples.
-Tools like proof assistants and logical frameworks are designed to support specifying and reasoning about dialects like these, and tools like compiler generators and language workbenches lower the engineering cost of implementing them.
+Tools like proof assistants and logical frameworks support specifying and reasoning metatheoretically about dialects like these, and tools like compiler generators and language workbenches lower the implementation cost, again contributing to their increasing proliferation.
 
@@
 
-The reason why this practice of using dialects as vehicles for new syntactic and semantic primitives is alarming is that it is, in an important sense,  anti-modular: a program written in one dialect cannot, in general, safely and idiomatically interface with libraries written in a different dialect.
+The reason why this proliferation of language dialects should be considered alarming is that it is, in an important sense,  anti-modular: a program written in one dialect cannot, in general, safely and idiomatically interface with libraries written in a different dialect.
 %At best, one can hope that the compilers for the decidedlytwo languages target a common intermediate language,
-This would fundamentally require combining the two dialects into a single language.
+To do so would fundamentally require combining the two dialects into a single language.
 Left unconstrained, this is an ill-posed problem, i.e.
 given two dialects that are specified by judgements of arbitrary form, it is not clear what it would even mean to combine them.
-Even when the dialects are specified using a formalism that does provide some way to na\"ively combine two languages, there is generally no guarantee that the composition will conserve important syntactic and semantic properties.
+Even when the dialects are specified using a formalism that does provide some way to na\"ively combine two dialects, there is generally no guarantee that the composition will conserve important syntactic and semantic properties.
 %There is no well-defined mechanism for constructing such a ``combined language'' in general.
@@
 
-Dialects can, however, have an indirect impact on large-scale software development because the language designers that control comparatively popular languages, like Ocaml and Scala, can occasionally be convinced to incorporate  ideas from dialects  into backwards compatible language revisions.
+Dialects can have an indirect impact on large-scale software development because the language designers that control comparatively popular languages, like Ocaml and Scala, can be convinced to incorporate  ideas from dialects  into backwards compatible language revisions.
 %These decisions are increasingly influenced by community processes, e.g.
@@
 %Dialects thus serve the role of rhetorical vehicles for new ideas, rather than direct artifacts. 
-These major languages have thus become more expressive over time.
-However, this has come at a cost: they have also ballooned in size.
-Moreover, because there appears to be a ``long tail'' of potentially useful primitives (for derived syntactic primitives, our group has gathered initial data speaking to this \cite{TSLs}),  primitives that are only situationally useful, or that trigger aesthetic disagreements between different factions of the community, are still often left languishing in ``toy'' dialects.
-A conservative approach to incorporating new primitives is sensible because once a primitive is introduced, it becomes entrenched as-is and monopolizes ``syntactic resources'', as we will discuss below.Recalling the words of  Reynolds, which are clearly as relevant today as they were nearly half a century ago \cite{Reynolds70}:%Because there is no data about how useful a construct is in practice until it is included in a language like this, decisions about which constructs to include are often informed by little more than intuition.
+These languages have, in this way, become more expressive over time.
+However, this expressiveness has come at a cost: they have also ballooned in size.
+Yet because there appears to be a ``long tail'' of potentially useful primitives (for derived syntax, our group has  gathered initial data speaking to this \cite{TSLs}),  primitives that are only situationally useful, or that trigger aesthetic disagreements between different factions of the community, are still often left languishing in ``toy'' dialects.
+Recalling the words of  Reynolds, which are as relevant today as they were nearly half a century ago \cite{Reynolds70}:%Because there is no data about how useful a construct is in practice until it is included in a language like this, decisions about which constructs to include are often informed by little more than intuition.
 %This approach is antithetical to the ideal of a truly \emph{general-purpose language} described at the beginning of this section.
@@
 and generality in language design is a serious unsolved 
-problem.}%\begin{flushright}-- John Reynolds (1970)\end{flushright}
+problem.}\begin{flushright}--- John Reynolds (1970)\end{flushright}
 \end{quote}
 
-\newpage There are two possible paths forward for the subset of the language design community interested in keeping general-purpose languages small and free of \emph{ad hoc} primitives.
-One, exemplified (arguably) by Standard ML, is to generally eschew the introduction of new primitives and settle on a set of primitives that sit at a ``sweet spot'' in the overall language design space, accepting that in some circumstances, these primitives trade away expressive power or have high syntactic cost.
+\newpage %A measured approach to incorporating new primitives into a general-purpose programming language is  sensible because once a primitive is introduced, it becomes entrenched as-is and monopolizes ``syntactic resources'', as we will discuss below.
+This leaves the language design community with two possible paths forward if we wish to keep general-purpose languages small and free of \emph{ad hoc} primitives.
+One, exemplified (arguably) by SML, is to generally eschew the introduction of new primitives and settle on a set of primitives that sit at a ``sweet spot'' in the overall language design space, accepting that in some circumstances, this trades away expressive power or leads to  high syntactic cost.
 % that this implies.
-The other path forward is to press on in the search for even more general language primitives, i.e.
-primitives that are so expressive that they allow us to degrade a broad class of existing primitives, including those found in dialects, into modularly composable library constructs, where they can be evaluated on their individual merits by programmers.
+The other path forward is to continue on in the search for even more general language primitives, i.e.
+primitives that are so expressive that they permit us to degrade a broad class of existing primitives, including those found in dialects, to modularly composable library constructs, where they can be evaluated on their individual merits by programmers and used together without the possibility of conflict.
 % This should render the construction of new dialects increasingly unnecessary.
-Such primitives do sometimes arise.
-For example, Ocaml recently introduced  ``generalized algebraic data types'' (GADTs), based on research on guarded recursive datatype constructors \cite{XiCheChe03}.
-Using GADTs, Ocaml was able to move some machinery for typechecking operations that use format strings, like \texttt{printf}, out of the language and into the standard library (though some machinery remains built in).
-Our broad aim will be to introduce a new general-purpose programming language called Verse\footnote{We distinguish Verse from Wyvern, which is the language referred to in prior publications about some of the work being proposed here, because Wyvern is a group effort evolving independently in some important ways.} that takes even more dramatic steps down this second path.
+Encouragingly, such new primitives do occasionally arise.
+For example, Ocaml recently added support for  ``generalized algebraic data types'' (GADTs), based on research on guarded recursive datatype constructors \cite{XiCheChe03}.
+Using GADTs, Ocaml was able to move some of the machinery for typechecking operations that use format strings, like \texttt{sprintf}, out of the language and into the standard library (unfortunately, some machinery remains built in).
+
+Our broad aim is to introduce a new functional programming language called Verse\footnote{We distinguish Verse from Wyvern, which is the language referred to in prior publications about some of the work being proposed here, because Wyvern is a group effort evolving independently in some important ways.} that takes even more dramatic steps down this second path.
 %Similarly, it recently introduced ``open datatypes'', which subsume its previous more specialized exception type, and captures many use cases for .
@@
 
+%\vspace{-10px}
 \section{Proposed Contributions}
 
-Verse has a module system taken directly from SML, but its core language is organized in a novel manner around a minimal typed lambda calculus, called the Verse \emph{internal language}, introduced in Section \ref{sec:verse}.
+Verse has a module system taken directly from SML, but its core language is organized in a novel manner around a minimal typed lambda calculus, called the Verse \emph{internal language}, discussed in Section \ref{sec:verse}.
 In lieu of typical {external} (i.e.
-programmer-facing) derived syntax and  type structure, Verse introduces two novel primitive constructs: \textbf{typed syntax macros} (TSMs),  in Section \ref{sec:syntax}, and \textbf{metamodules},  in Section \ref{sec:metamodules}.
-Briefly summarized, TSMs subsume the need for derived syntax specific to library constructs (e.g.
-list syntax in SML) by giving library providers static control over the parsing and elaboration of delimited segments of textual concrete syntax.
-Metamodules give library providers static hooks directly into the semantics of the Verse EL, allowing  library providers to, for example, define the type structure of records (and various ``record-like'' constructs, as discussed above) by type-directed translation to the internal language (which builds in only nullary and binary products).
-Both TSMs and metamodules can be understood as novel \emph{metaprogramming} primitives, because they involve static code generation.
-We will  also introduce a simple variant of each of these primitives that leverages Verse's support for local type inference to further reduce syntactic cost in certain common situations.
+programmer-facing) derived syntax and  type structure, Verse introduces two novel primitive constructs:
+\begin{itemize}
+\item \textbf{Typed syntax macros} (TSMs), which we introduce in Section \ref{sec:syntax}, subsume the need for derived syntax specific to library constructs, e.g.
+list syntax as built in to SML, by giving library providers static control over the parsing and elaboration of delimited segments of concrete syntax.
+\item \textbf{Metamodules}, which we introduce in Section \ref{sec:metamodules}, give library providers static hooks directly into the semantics of the Verse external language, allowing  them to define, for example, the type structure of records (and various ``record-like'' constructs, as discussed above) by type-directed translation to the more minimal internal language (which builds in nullary and binary products, but not records).
+\end{itemize} 
+Both TSMs and metamodules can be understood as \emph{metaprogramming} primitives, because they involve static code generation.
+%We will  also introduce a simple variant of each of these primitives that leverages Verse's support for local type inference to further reduce syntactic cost in certain common situations.
 
 The key challenge in the design of these primitives will come in ensuring that they are metatheoretically well-behaved, given that they aim to give library providers decentralized control over aspects of the language that have typically been under the monolithic control of the language designer.
-If we are not careful, many of the problems mentioned above as inherent to combining distinct language dialects could simply manifest themselves in the semantics of these primitives.\footnote{This is why languages  like Verse are often called \emph{extensible languages}, though this is somewhat of a misnomer.
+If we are not careful, many of the problems discussed above as inherent to combining distinct language dialects could simply manifest themselves in the semantics of these primitives.\footnote{This is why languages  like Verse are often called \emph{extensible languages}, though this is somewhat of a misnomer.
 The chief characteristic of an extensible language is that it \emph{doesn't} need to be extended in situations where other languages would need to be extended.
 We will avoid this somewhat confusing terminology.} Our main technical contributions will come in showing how to address these problems in a principled manner.
-In particular, syntactic ambiguities will be impossible by construction and  we will validate the code  statically generated by TSMs and metamodules to maintain a hygienic type discipline and, most uniquely, powerful modular reasoning principles.
-Library providers will be able to reason about the constructs that they have defined in isolation, and library clients will be able to use them safely in any combination, without the possibility of conflict.\footnote{We will assume that simple naming conflicts can be avoided extrinsically in some suitable manner, e.g.
+In particular, syntactic conflicts will be impossible by construction and  we will validate the code  statically generated by TSMs and metamodules to maintain a hygienic type discipline and, most uniquely, powerful modular reasoning principles -- library providers will be able to reason about the constructs that they have defined in isolation, and library clients will be able to use them safely in any combination, without the possibility of conflict.\footnote{We  assume that simple naming conflicts can be avoided extrinsically, e.g.
 by using a URI-based naming scheme as in the Java ecosystem.}
@@
 \begin{quote}
-A programming language can  give library providers the power to meta\-pro\-gram\-matic\-ally express new derived syntax and external type structure atop a minimal typed internal language, using  primitives that maintain a hygienic type discipline and modular reasoning principles.
-These  primitives are  expressive enough to subsume the need for a variety of primitives that are, or would need to be, built in to comparable contemporary languages.
+A functional programming language can give library providers the ability to meta\-pro\-gram\-matic\-ally express new derived syntax and external type structure atop a minimal typed internal language while maintaining a hygienic type discipline and modular reasoning principles.
+%These  primitives are  expressive enough to subsume the need for a variety of primitives that are, or would need to be, built in to comparable contemporary languages.
 \end{quote}
@@
 It may also be useful to explicitly acknowledge that library providers could leverage the primitives we introduce   to define constructs that are in rather poor taste.
-We  expect that in practice, Verse will come with a standard library defining a carefully curated collection of standard constructs, as well as guidelines for advanced users regarding when it would be sensible to use the primitives we introduce (following the example of languages that support operator overloading or typeclasses, which also have the potential for ``abuse'').
+We  expect that in practice, Verse will come with a standard library defining a carefully curated collection of standard constructs, as well as guidelines for advanced users regarding when it would be sensible to use the mechanisms we introduce (following the example of languages that support operator overloading or typeclasses, which also have the potential for ``abuse'').
 %The vast majority of programmers should not use the primitives that we introduce directly.
@@
 Finally, we are not interested here in languages that feature full-spectrum dependent types, which blur the phase separation between compile-time and run-time, though we conjecture that the primitives we introduce could be introduced into languages like Gallina (the ``external language'' of the Coq proof assistant) with some modifications.
-Verse should be compared to languages that maintain a phase separation, like ML and Scala.
+Verse should be compared to languages that maintain a phase separation, like ML, Haskell and Scala.
 %Our interest is in making sure that the standard library is no more privileged than any other library.
@@
 
-\section{Verse}\label{sec:verse}
-Let us begin by briefly summarizing how Verse is organized and specified.
-Verse consists of a \emph{module language} and a \emph{core language}.
-The module language is based directly on SML's module language,  which we assume familiarity with for the purposes of this work \cite{harper1997programming,MacQueen:1984:MSM:800055.802036}.
-We will give an example of its use in Section \ref{sec:syntax}, but do not plan to specify it in detail.
+\section{Summary of Judgements}\label{sec:verse}
+To situate ourselves within a formal framework, let us begin with a brief summary of how Verse is organized and specified.
+Verse consists of a \emph{module language} atop a \emph{core language}.
+The module language is based directly on SML's module language,  which we assume working familiarity with for the purposes of this proposal \cite{harper1997programming,MacQueen:1984:MSM:800055.802036}.
+We will give examples of its use in both Section \ref{sec:syntax} and Section \ref{sec:metamodules}, but do not plan to specify it in detail.
 
 The core language is organized much like the first stage of a type-directed compiler (e.g.
-the TIL compiler for Standard ML \cite{tarditi+:til-OLD}), consisting of a user-facing \emph{typed external language} (EL) specified by type-directed translation to a minimal \emph{typed internal language} (IL).
-The main judgements in the specification of the EL take the following form (we omit mention of various contexts for now):
+the TIL compiler for Standard ML \cite{tarditi+:til-OLD}), consisting of a user-facing \emph{typed external language} (EL) specified by type-directed translation to a minimal \emph{typed internal language} (IL).\footnote{Note that we will use the terms ``expression'' and ``type'' without qualification to refer to  external expressions and types, respectively.
+} The main judgements in the specification of the EL take the following form (omitting various contexts for now):
 \\[1ex]
 $\begin{array}{ll}
-\textbf{Judgement} & \textbf{Pronunciation}\\
-e \Rightarrow \sigma \leadsto \iota & \text{External expression $e$ synthesizes external type $\sigma$ and has translation $\iota$.}\\
-e \Leftarrow \sigma \leadsto \iota & \text{External expression $e$ analyzes against external type $\sigma$ and has translation $\iota$.}\\
-\sigma~\mathtt{type} \leadsto \tau & \text{Static expression $\sigma$ is an external type with translation $\tau$.}
+\textbf{Judgement Form} & \textbf{Pronunciation}\\
+\vdash e \Rightarrow \sigma \leadsto \iota & \text{Expression $e$ synthesizes type $\sigma$ and has translation $\iota$.}\\
+\vdash e \Leftarrow \sigma \leadsto \iota & \text{Expression $e$ analyzes against type $\sigma$ and has translation $\iota$.}\\
+\vdash \sigma~\mathtt{type} \leadsto \tau & \text{Static expression $\sigma$ is a type with translation $\tau$.}
 \end{array}
 $\\[1ex]
-Note that the typing judgements are \emph{bidirectional}, i.e.
-there is a distinction between \emph{type synthesis} (the type is an ``output'') and \emph{type analysis} (the type is an ``input'') \cite{Pierce:2000:LTI:345099.345100}.
-This is to  allow us to explicitly specify how Verse's \emph{local type inference} works, and in particular, how it interacts with the primitives that we will introduce.
-Like Scala, we intentionally do not aim to support global type inference, both because it is not compatible with the primitives we introduce as described and because we believe that in practice, the cognitive cost of the error messages that arise outweigh its marginal utility relative to local type inference.
-Note also that, for concision, we will write ``expression'' and ``type'' without qualification to refer to  external expressions and types, respectively.
+Note that the expression typing judgements are \emph{bidirectional}, i.e.
+we make a judgemental distinction between \emph{type synthesis} (the type is an ``output'') and \emph{type analysis} (the type is an ``input'') \cite{Pierce:2000:LTI:345099.345100}.
+This is to  allow us to explicitly specify how Verse's \emph{local type inference} works, and in particular, how it interacts with the primitives that we aim to introduce.
+Like Scala, we intentionally do not attempt to support global type inference, primarily because it is not technically possible to do so (but also because we believe that compilers for languages that support only local type inference generate clearer error messages).
 
@@
 
-Our design is largely insensitive to the details of the IL.
-The only strict requirements are that the IL is type safe and supports parametric type abstraction.
+Our design is largely insensitive to the details of the internal language -- the only strict requirements are that the IL be type safe and support parametric type abstraction.
 For our purposes, we will keep things simple by using the strictly evaluated polymorphic lambda calculus with nullary and binary product and sum types and recursive types, which we assume the reader has familiarity with and for which all the necessary metatheorems are well-established (we follow \cite{pfpl}).
-The main judgements in the static semantics of the IL (again omitting contexts for now) thus take the following form:
+The main judgements in the static semantics of the IL (again omitting contexts for now) take the following essentially standard form:
 \\[1ex]
@@
 \begin{array}{ll}
-\textbf{Judgement} & \textbf{Pronunciation}\\
-\iota : \tau & \text{Internal expression $\iota$ has internal type $\tau$.}\\
-\tau~\mathtt{type} & \text{Internal type $\tau$ is valid.}
+\textbf{Judgement Form} & \textbf{Pronunciation}\\
+\vdash \iota : \tau & \text{Internal expression $\iota$ has internal type $\tau$.}\\
+\vdash \tau~\mathtt{itype} & \text{Internal type $\tau$ is valid.}
 \end{array}
@@
 \noindent
-The dynamic semantics can be specified as a transition system with  judgements of the following form:
+The dynamic semantics can be specified also in the standard manner as a transition system with  judgements of the following form:
 \\[1ex]
@@
 \begin{array}{ll}
-\textbf{Judgement} & \textbf{Pronunciation}\\
+\textbf{Judgement Form} & \textbf{Pronunciation}\\
 \iota \mapsto \iota' & \text{Internal expression $\iota$ transitions to $\iota'$.}\\
@@
 
-Primitive features like state, exceptions, concurrency primitives, scalars, arrays and others characteristic of a first-stage compiler IL would also be included in practice, and in some cases this would affect the shape of the internal semantics, but we omit them when their inclusion would not meaningfully affect our discussion.
+Features like state, exceptions, simple concurrency primitives, scalars, arrays and others characteristic of a first-stage compiler intermediate language would also be included in the IL in practice, and in some cases this would affect the shape of the internal semantics, but we omit them for simplicity.
 %Features like record types and labeled sum types, on the other hand, are not built in.
@@
 newlines and indentation are not ignored).
-This design choice is not  fundamental to our proposed contributions, but it will be useful for cleanly expressing a class of examples that we will discuss later.
-We will specify key aspects of its concrete syntax with an \emph{Adams grammar} \cite{Adams:2013:PPI:2429069.2429129} (such a specification for Wyvern, which has a very similar syntax, can be found in \cite{TSLs}), but for the purposes of this proposal, we will simply introduce Verse's concrete syntax by example as we go on.
+This design choice is not  fundamental to our proposed contributions, but it will be useful for cleanly expressing a class of examples that we plan to discuss later.
+We plan to specify some novel aspects of Verse's concrete syntax with an \emph{Adams grammar} \cite{Adams:2013:PPI:2429069.2429129} (such a specification for Wyvern, which has a very similar syntax, can be found in \cite{TSLs}), but for the purposes of this proposal, we will simply introduce Verse's concrete syntax by example as we go on.
+%For constructs that have an obvious analog in ML, we will omit a detailed explanation.
 
@@
 Many languages go further, building in syntax associated with various other types of data, like vectors (SML/NJ), arrays (Ocaml), commands (Haskell), syntax trees (Scala), XML data (Scala, Ur/Web) and SQL queries (F\#, Ur/Web).
+%This is a rather \emph{ad hoc} process.% discussed previously, the usual approach is to require that the language designer build in new derived syntactic forms.
 %The desugaring from the latter to the former is specified by the language itself.
@@
 %For forms with a clear analogy to a form in Standard ML, we will assume the  semantics are analagous without providing details.
-To motivate our desire for this level of syntactic control, we begin in Sec.
+To specifically motivate our desire for this level of syntactic control, we begin in Sec.
 \ref{sec:examples} with a representative example: regular expression patterns expressed using abstract data types.
-We show how the usual workaround of using strings to introduce patterns is not ideal.
+We show how the usual approach of using strings to introduce patterns is not ideal.
 We then survey existing alternatives in Sec.
@@
 In Sec.
-\ref{sec:syntax-contributions}, we outline our proposed mechanisms and discuss how they resolve these issues.
+\ref{sec:syntax-contributions}, we outline our proposed mechanisms and discuss how they will resolve these issues.
 We conclude in Sec.
@@
 \subsection{Motivating Example: Regular Expression Syntax}\label{sec:examples}
-Let us begin by taking the perspective of a \emph{regular expression} library provider.
+Let us begin from the perspective of a regular expression library provider.
 Recall that regular expressions are a common way to capture patterns in strings \cite{Thompson:1968:PTR:363347.363387}.
@@
 However, there are some reasons not to expose this representation of patterns directly to clients.
-First, patterns are usually identified up to their reduction to a normal form.
+First, such patterns are usually identified up to their reduction to a normal form.
 For example, $\textbf{seq}(\textbf{empty}, p)$ has normal form $p$.
@@
 a corresponding finite automata) without exposing this ``implementation detail'' to clients.
-Indeed, there are many ways to represent regular expression patterns, each with different performance trade-offs.
-For these reasons, a better approach is to use \emph{abstract data types}.
-In Verse, as in ML, this is a mode of use of the module system.
+Indeed, there may be many ways to represent regular expression patterns, each with different performance trade-offs.
+For these reasons, a better approach is to use \emph{abstract data types}, which in Verse, as in ML, are a mode of use of the module system.
 In particular, we can define the following \emph{module signature}, where the type of patterns, \lstinline{t}, is held abstract.
@@
 To express the concrete syntax of patterns, regular expression library providers usually provide a utility function that transforms strings to patterns.
-As just mentioned, there  may be many implementations of the \lstinline{PATTERN} signature, so the standard approach is to define a \emph{parameterized module} (a.k.a.
+Because, as just mentioned, there  may be many implementations of the \lstinline{PATTERN} signature, the standard approach is to define a \emph{parameterized module} (a.k.a.
 \emph{functor} in SML) defining utility functions generically:
@@
  %Again, none of our goals are comprehensively achieved.
-This approach is imperfect for several reasons:
+Unfortunately, this approach is imperfect for several reasons:
 \begin{enumerate} 
-\item Our first problem is syntactic: string escape sequences conflict syntactically with pattern escape sequences. For example, the following will not be well-formed:
+\item First, there are syntactic conflicts between string escape sequences and pattern escape sequences. For example, the following is not a well-formed term:
 \begin{lstlisting}[numbers=none,mathescape=|]
@@
 \end{lstlisting}
-When compiling an analagous expression using \verb|javac|, we encounter the error message \verb|error: illegal escape character|.\footnote{When compiling an analagous expression using SML of New Jersey (SML/NJ), we encounter the seemingly inexplicable error message \texttt{Error: unclosed string}.} In a small lab study, we observed that this error was common, and nearly always initially misinterpreted by even experienced programmers who hadn't used regular expressions recently \cite{Omar:2012:ACC:2337223.2337324}.
+When compiling an expression like this, the programmer would see an error message like \verb|error: illegal escape character|.\footnote{This is the error message that \texttt{javac} produces.
+When compiling an analagous expression using SML of New Jersey (SML/NJ), we encounter the rather bizarre error message \texttt{Error: unclosed string}.} In a small lab study, we observed that this sort of error was common, and that solving the problem was nearly always initially challenging to even experienced programmers who hadn't used regular expressions recently \cite{Omar:2012:ACC:2337223.2337324}.
 The inelegant workaround is to  double backslashes:
@@
 
-Some languages, anticipating such modes of use, build in special string forms that leave escape sequences uninterpretted. For example, Ocaml allows the following, which is perhaps more acceptable:
+Some languages, anticipating such modes of use, build in alternative string forms that leave escape sequences uninterpretted. For example, Ocaml allows the following, which is perhaps more acceptable:
 \begin{lstlisting}[numbers=none]
@@
 
-\item The final problem is that using strings to construct patterns encourages programmers to use string concatenation inappropriately to construct patterns derived from user input.
+\item The final problem is that using strings to construct patterns encourages programmers to use  operations over strings, like concatenation, to construct patterns derived from user input that happens also to be of type \lstinline{string}.
+This can lead to subtle but serious bugs.
 For example, consider the following function:
@@
 \end{lstlisting}
-The (unstated) intent here is to treat \lstinline{name} as a sub-pattern matching only itself, but this is not the observed behavior when \lstinline{name} contains special characters that mean something else in patterns.
+The (unstated) intent here is to treat \lstinline{name} as a sub-pattern matching only itself, but this is not the observed behavior when \lstinline{name} contains special characters that have other meanings in patterns.
 The correct code again resembles abstract syntax:
@@
 \end{lstlisting}
-
-In applications that query sensitive data, failures of this variety can lead to \emph{injection attacks}.
-These are observed to be both common and catastrophic from the perspective of security \cite{owasp2013}.
-The attack vector above was, of course, the fault of the programmer using a flawed heuristic, so it could have been avoided with sufficient discipline.
-The problem is again that it is difficult to detect violations of this discipline automatically.
-Both functions above have the same type and behave identically at many inputs, particularly those that would be expected during typical executions of the program (i.e.
-alphabetic names), so testing often misses the problem.
-Formally proving that such problems do not arise involves reasoning about complex run-time data flows.
+Note that both functions have the same type and behave identically at many inputs, particularly those that would be expected during typical executions of the program (i.e.
+alphabetic names), so testing can miss the problem.
+Proving that such problems cannot arise involves reasoning about complex run-time data flows, so it is once again difficult to automate.
+In applications that query sensitive data, mistakes like this lead to \emph{injection attacks}, which are both common and catastrophic from the perspective of software security \cite{owasp2013}.
+%Ultimately, of course, mistakes like this are the fault of a programmer using a flawed heuristic, and they could be avoided with discipline.
+The problem is once again that it is difficult to detect violations of this discipline automatically.
 
@@
 \end{enumerate}
+
 % \begin{enumerate}
@@
 
-\subsection{Existing Approaches}\label{sec:syntax-existing} These difficulties with string-oriented approaches to concrete syntax are not particular to regular expression patterns, but rather arise in many similar scenarios, motivating much research on reducing the need for run-time string parsing \cite{Bravenboer:2007:PIA:1289971.1289975}.
-Existing approaches take one of two approaches: \emph{syntax extension} or \emph{static term rewriting}.
-
-\subsubsection{Syntax Extension}
-One common approach is to use a system that allows users to directly extend the syntax of a language with new derived syntax, in this case for patterns.
+	\subsection{Existing Approaches}\label{sec:syntax-existing} The difficulties above are common whenever one uses  strings transiently to construct more richly structured data, merely for the sake of syntactic convenience.
+%Strings are, simply put, not ideally suited for this task.
+	This key observation has motivated a substantial quantity of research on reducing the need for run-time string parsing \cite{Bravenboer:2007:PIA:1289971.1289975}.
+Existing approaches can be classified in  one of two ways: as being based on \emph{syntax extension} or \emph{static term rewriting}.
+
+\subsubsection{Syntax Extension}\label{sec:syntax-extension}
+One approach to address the problems above would be to leverage a system that allows us to directly extend the syntax of our language with new derived syntactic forms.% for patterns.
 
 The simplest such systems are those where each new syntactic form is described by a single rewrite rule.
-For example, the external language of Coq (called Gallina) includes such a system \cite{Coq:manual}.
-A formal account of such systems has been developed by Griffin \cite{5134}.
-Unfortunately, these systems are not able to express pattern syntax in the conventional manner.
+For example, Gallina, the external language of the Coq proof assistant, provides such a system \cite{Coq:manual}.
+A formal account of systems like these has been developed by Griffin \cite{5134}.
+Unfortunately, these systems are not flexible enough to allow us to express pattern syntax in the conventional manner.
 For example, sequences of characters  can only be parsed as identifiers using these systems, rather than as characters in a pattern.
-Syntax extension systems based on context-free grammars like  Sugar* \cite{erdweg2013framework}, Camlp4 \cite{ocaml-manual} and many others are more expressive, and would allow us to directly introduce pattern syntax into our core language's grammar.
-
-However, this is a perilous approach because none of the systems described thusfar guarantee \emph{syntactic composability}, i.e.
+Syntax extension systems based on context-free grammars like  Sugar* \cite{erdweg2013framework}, Camlp4 \cite{ocaml-manual} and many others are more expressive, however, and thus would allow us to directly introduce pattern syntax into our core language's grammar, perhaps like this:
+\begin{lstlisting}[numbers=none]
+let ssn = /\d\d\d-\d\d-\d\d\d\d/
+\end{lstlisting}
+
+However, this is a rather perilous approach because none of these systems guarantee freedom from \emph{syntactic conflicts}, i.e.
 as the Coq manual states: ``mixing different symbolic notations in [the] same text may cause serious parsing ambiguity''.
-If another library provider used similar syntax for a different variant of regular expressions, or for an entirely unrelated abstraction, then a client could not simultaneously use both libraries in the same scope.
-Properly considered, every combination of extensions introduced by one of these mechanisms creates a \emph{de facto} derived syntactic dialect of the language.
+So if another library provider used similar syntax for a different implementation of regular expressions, or for some other unrelated abstraction, then a client could not simultaneously use both libraries in the same scope.
+Properly considered, every combination of extensions introduced by one of these mechanisms creates a \emph{de facto} derived syntactic dialect of our language.
 % Resolving such parsing amibiguities is left to each client of the library.
 
-In response to this problem, Schwerdfeger and Van Wyk have developed a modular analysis that accepts only grammar extensions that use a unique starting token and satisfy some subtle conditions on follow sets of base language non-terminals \cite{conf/pldi/SchwerdfegerW09}.
-Such extensions can be used together in any combination.
-However, simple starting tokens like \lstinline{pattern} cannot be guaranteed to be globally unique, so we would need to use a more verbose starting token like \lstinline{edu_cmu_verse_rx_pattern}.
-There is no principled way to define local abbreviations for starting tokens because this mechanism is language-external.
+In response to this problem, Schwerdfeger and Van Wyk developed a modular analysis that accepts only grammar extensions that specify a universally unique starting token and obey subtle constraints on  the follow sets of base language non-terminals \cite{conf/pldi/SchwerdfegerW09}.
+Extensions that satisfy these criteria can be used together in any combination without the possibility of conflict.
+However, note that the most natural starting tokens like \lstinline{pattern} cannot be guaranteed to be universally unique, so we would be forced to use a more verbose token like \lstinline{edu_cmu_verse_rx_pattern}.
+There is no principled way for clients of our extension to define local abbreviations for starting tokens because this mechanism is language-external.
 
 Putting this issue aside, we must now decide how our newly introduced derived forms elaborate to forms in our base grammar.
-In this case, because we are using a modular encoding of patterns, we must first decide which module the desugaring will use.
-Clearly, simply assuming that a module identified as \lstinline{P} satisying \lstinline{PATTERN} is in scope is a brittle solution.
+This is tricky because we have defined a modular encoding of patterns -- which particular module should the elaboration logic use?
+Clearly, simply assuming that some module identified as \lstinline{P} satisying \lstinline{PATTERN} is in scope is a brittle solution.
 Indeed, we should expect that the extension mechanism actively prevents such capture of specific variable names to ensure that variables (including module variables) can be freely renamed.
@@
 
-Putting aside the question of hygiene as well, we can address the problem by requiring that the client explicitly identify the module the desugaring should use:
+Putting aside this question of hygiene as well, we can address the problem by requiring that the client explicitly identify the module the elaboration should use:
 \begin{lstlisting}[numbers=none]
@@
 \end{lstlisting}
-For patterns constructed compositionally, we can define \emph{splicing} syntax for both strings and other patterns:
+For patterns constructed compositionally, we can define \emph{splicing} syntax for both strings and other patterns, e.g. as shown below:
 \begin{lstlisting}[numbers=none,escapechar=|]
@@
 \end{lstlisting}
-The body of this function elaborates to the body of \lstinline{example_fixed} as shown above.
-Had we mistakenly used the pattern splicing prefix, \lstinline{%name}, rather than the string splicing prefix, \lstinline{@name}, we would get a static type error, rather than a silent security vulnerability.
-
-Such syntax extension approaches suffer from a paucity of direct reasoning principles.
-Given an unfamiliar piece of syntax, there is no simple method for determining what type it will have, or even for identifying which extension determines its desugaring, causing difficulties for both humans (related to code comprehension) and tools.
-
-\subsubsection{Term Rewriting}
+%The body of this function elaborates to the body of \lstinline{example_fixed} as shown above. 
+Had we mistakenly used the pattern splicing syntax, writing \lstinline{%name}, rather than the string splicing syntax, \lstinline{@name}, we would get a static type error, rather than the  silent injection  vulnerability that we discussed earlier.
+
+A general problem with approaches like these (in addition to those already mentioned) is that they  suffer from a paucity of direct reasoning principles -- given an unfamiliar piece of syntax, there is no simple method for determining what type it will have, or even for identifying which extension determines its elaboration, causing difficulties for both humans (related to code comprehension) and tools.
+
+\subsubsection{Static Term Rewriting}
 An alternative approach is to leave the concrete syntax of the language fixed, but repurpose it for novel ends using a \emph{local term-rewriting system}.
-Among the earliest examples is the LISP macro system \cite{Hart63a}.
-This system was later refined, notably in Scheme and its dialects to better support reasoning compositionally about the rewriting that is performed \cite{Kohlbecker86a}.
+Among the earliest examples is the LISP macro system \cite{Hart63a}, later modified, notably in the Scheme language, to support reasoning hygienically about the rewriting that is performed \cite{Kohlbecker86a}.
 In languages with a stronger static type discipline, variations on macros that restrict rewriting to a particular type and perform the rewriting in an explicitly staged manner have also been studied \cite{Herman10:Theory,ganz2001macros} and integrated into languages, e.g.
@@
 Here, \lstinline{pattern} is a macro parameterized by a module \lstinline{P : PATTERN}.
-It statically parses the provided string literal (which must still be written using an Ocaml-style literal here) to generate an elaboration, at type \lstinline{P.t}.
-
-For patterns that are constructed compositionally, we need to get more creative. For example, we might repurpose the infix operators that the language provides for other purposes:
+It statically parses the provided string literal (which must still be written using an Ocaml-style literal here) to generate an elaboration, specified by the macro to be at type \lstinline{P.t}.
+
+For patterns that are constructed compositionally, we need to get more creative.
+For example, we might repurpose the infix operators that the language normally uses for other purposes to support string and pattern splicing as follows:
 
@@
 
-This does not give us syntax that is quite as clean and clear as the syntax available using syntax extensions, but it does address a number of its deficiencies: there cannot be syntactic conflicts (because the syntax is not extended), we can define macro abbreviations because macros are integrated into the language, there is a hygiene discipline that guarantees that the elaboration will not capture variables inadvertently, and there is a typing discipline (i.e.
-we need not determine the elaboration to know what type it will have).
+While this does not leave us with syntax that is quite as clean as would be possible with a  na\"ive syntax extension,  it does address many of the problems we discussed above: there cannot be syntactic conflicts (because the syntax is not extended), we can define macro abbreviations because macros are integrated into the language, there is a hygiene discipline that guarantees that the elaboration will not capture variables inadvertently, and by using a typed macro system, there is a typing discipline (i.e.
+creativewe need not examine the elaboration directly to know what type it must have).
 
@@
 \subsection{Contributions}\label{sec:syntax-contributions}
-We propose a new primitive, the \textbf{typed syntax macro} (TSM), that combines the flexibility of syntax extensions with the reasoning guarantees of statically-typed macros, and show how to integrate it into a language with an ML-style module system, addressing all of the problems described above.
-We then introduce \textbf{type-specific languages} (TSLs), which leverage local type inference to control TSM dispatch.
-This further decreases syntactic cost (which is, of course, the entire purpose of this exercise).
-The result is that derived syntax defined using TSLs  approaches the cost of derived syntax built in primitively.
+We propose a new primitive, the \textbf{typed syntax macro} (TSM), that combines the flexibility of syntax extensions with the reasoning guarantees of statically-typed macros, and show how to integrate it into a language with an ML-style module system, addressing all of the problems described thusfar.
+We then introduce \textbf{type-specific languages} (TSLs), which leverage local type inference to control TSM dispatch, further decreasing syntactic cost (which is, of course, the entire purpose of this exercise).
+The result is that derived syntax defined by library providers using TSMs and TSLs  approaches the cost of derived syntax built in primitively by a language designer (using a na\"ive tool like Camlp4).
 
@@
 %A typed syntax macro is invoked by applying it to a \emph{delimited form}, which can contain  arbitrary syntax in its \emph{body}.  
-As our first example, consider the following concrete term:
+Consider the following concrete external expression:
 \begin{lstlisting}[numbers=none]
@@
 \end{lstlisting}
-Here, a parameterized TSM \lstinline{pattern} is applied to a module parameter, \lstinline{P}, and a {delimited form}, \lstinline{/SURLA|T|G|CEURL/}.
-Based on the body of the delimited form, i.e.
-the characters between the delimiters, the TSM statically computes an elaboration to the following {base external form}:
+Here, a parameterized TSM, \lstinline{pattern}, is applied to a module parameter, \lstinline{P}, and a {delimited form}, \lstinline{/SURLA|T|G|CEURL/}.
+The TSM statically parses the body of the provided delimited form, i.e.
+the characters between the delimiters (shown here in blue), and computes an elaboration, i.e.
+another external expression.
+In this case, it is as if we wrote the following directly:
 
@@
 \end{lstlisting}
-The parameterized TSM \lstinline{pattern} is defined as shown below:
+
+%The original expression, above, is statically rewritten to this expression.
+The definition of \lstinline{pattern} is shown below:
 \begin{lstlisting}[numbers=none]
@@
 \end{lstlisting}
-We declare a module parameter (calling it \lstinline{P} here is unimportant, i.e.
-the TSM can be used with \emph{any} module satisfying module type \lstinline{PATTERN}), which is bound in the scope of the type clause \lstinline{at P.t}.
-This clause  guarantees that all elaborations successfully generated by this TSM will be of type \lstinline{P.t}.
+Note that identifying the module parameter as \lstinline{P} here is unimportant, i.e.
+the TSM can be used with \emph{any} module matching signature \lstinline{PATTERN}.
+The module parameter is bound in the scope of the type clause \lstinline{at P.t}.
+This clause  specifies that all elaborations successfully generated by this TSM will be of type \lstinline{P.t}.
 Elaborations are generated by the static action of the parse function defined in the indented block on the next line.
-The parse function must be of type \lstinline{ParseStream -> Exp}, where the type \lstinline{ParseStream} gives the function access to the \emph{body} of the delimited form (in blue above) and the type \lstinline{Exp} is a case type that encodes the abstract syntax of the base language (i.e.
+The parse function must be of type \lstinline{ParseStream -> Exp}, where the type \lstinline{ParseStream} gives the function access to the {body} of the delimited form (in blue above) and the type \lstinline{Exp} is a case type that encodes the abstract syntax of the base language (i.e.
 the Verse EL).
@@
 
-In positions where the type is known, e.g. due to a type ascription or at the argument position of a function call, the module parameter \lstinline{P} can be inferred:
+For concision in analytic positions, i.e. those where an expected type is known due to, for example, a type ascription, the module parameter \lstinline{P} can be inferred:
 \begin{lstlisting}[numbers=none]
@@
 
-TSMs can treat portions of the body of the delimited form as a spliced base language term, to support splicing syntax:
+TSMs can identify portions of the body of the delimited form that should be treated as spliced expressions, allowing us to support splicing syntax as described in Sec. \ref{sec:syntax-extension}:
 \begin{lstlisting}[numbers=none]
@@
 \end{lstlisting}
-A hygiene mechanism ensures that only those portions of the elaboration derived from such spliced terms can refer to variables in the surrounding scope, preventing inadvertent variable capture by the elaboration.
-In other words, the generated elaboration must be context-independent.
-
-\paragraph{Formalization} To give a flavor of the formal specification underlying TSMs, let us look at the rule for handling TSM invocation in a synthetic position. 
+The hygiene mechanism ensures that only those portions of the elaboration derived from such spliced expressions can refer to variables at the site of use, preventing inadvertent variable capture by the elaboration.
+Put another way, the generated elaboration must be context-independent.
+
+\paragraph{Formalization} To give a flavor of the formal specification underlying TSMs, let us look at the rule for handling TSM invocation in a synthetic position: 
 \begin{mathpar}
@@
 \end{mathpar}
-Note that the judgement forms are  slightly simplified in that they omit various contexts relevant only to other Verse primitives.
+Note that the judgement forms are  slightly simplified here, e.g.
+they omit various contexts relevant only to other Verse primitives.
 We include only \emph{typing contexts}, $\Gamma$, which map variables to types, and \emph{type abstraction contexts}, $\Delta$, which track universally quantified type variables (both in the standard way).
-The premises have the following purposes:
+The premises have the following meanings:
 \begin{enumerate}
 \item The first premise can be pronounced ``Under $\Delta$, TSM expression $s$ is a TSM at type $\sigma$ with parser $\iota_\text{parser}$''.
-Application of a parameterized TSM to its parameters is handled by this judgement (not shown).
+We elide the details of issues like module parameter application here for simplicity.
+% Application of a parameterized TSM to its parameters is handled by this judgement (not shown).
 \item The second premise creates a parse stream, $\iota_\text{ps}$, from the body of the delimited form.
@@
 \item The fourth premise decodes $\iota_\text{elab}$, producing the elaboration $e_\text{elab}$.
-\item The fifth premise validates the elaboration by analyzing it against the type $\sigma$ under an empty context.
-The \emph{current} typing and kinding contexts are ``saved'' for use when a spliced term is encountered during this process by setting them as the new \emph{outer contexts}.
+\item The fifth premise validates the elaboration by analyzing it against the type $\sigma$ under empty contexts.
+The \emph{current} typing and type abstraction contexts are ``saved'' for use when a spliced term is encountered during this process by setting them as the new \emph{outer contexts}.
 
@@
 \end{mathpar}
-For example, the elaboration generated in \lstinline{example_tsm} above would be:
+For example, the elaboration generated in \lstinline{example_tsm} above would, if written concretely, be:
 \begin{lstlisting}[numbers=none]
@@
 
-We will give a more complete account of all of these judgements later.
+%We plan to give a more complete account of all of these judgements in the dissertation.
 \subsubsection{Type-Specific Languages (TSLs)}
@@
 
- Local type inference then determines which TSM is applied when analyzing a delimited form not prefixed by a TSM name. For example, the following is equivalent to the above:
+ Local type inference then determines which TSM is applied when analyzing a delimited form not prefixed by a TSM name. For example, this is equivalent to \lstinline{example_tsm}:
 \begin{lstlisting}[numbers=none]
@@
 \end{lstlisting}
-
+From the client's perspective, it is essentially as if the language had built in derived syntax for lists and regular expression patterns directly.
+%The only constraint is that this syntax must be used in an analytic position, which we argue is actually better for code compren when encountering unfamiliar syntax.
 \subsection{Timeline}\label{sec:syntax-timeline}
-We have described and formally specified TSMs in a recently published paper \cite{sac15}, and TSLs in a paper last year \cite{TSLs}, both in the context of the Wyvern language.
-In the context of Verse, there are some changes that need to be made in the formalization.
-In particular, neither paper covered integration with an ML-style module system.
+We have described and given a more detailed formal specification of TSMs in a recently published paper \cite{sac15}, and TSLs in a paper last year \cite{TSLs}, both in the context of the Wyvern language.
+At the time of publication, Wyvern did not specify a module system, so in the context of Verse, there are some changes that need to be made in the formalization to handle module parameters.
 We plan to complete this in the course of writing the dissertation.
@@
 \section{Modularly Metaprogrammable Type Structure}\label{sec:metamodules}
-The Verse EL also does not build in substantial type structure.
-Only functions are built in directly.
-The type structure of case types (shown in use above), tuples, records, objects and other more esoteric constructs that we will describe later can all be expressed as modes of use of \emph{metamodules} (which are distinct from, though conceptually related to, \emph{modules}).
-
-The remainder of this section is organized like Section \ref{sec:syntax}.
-We begin with an example of type structure that cannot be expressed in languages like ML directly in Sec.
-\ref{sec:metamodules-example}, then show how existing approaches are insufficient in Sec.
-\ref{sec:metamodules-related}.
-Then we outline our approach in Sec.
-\ref{sec:metamodules-solution} and conclude with a timeline for remaining work in Sec.
-\ref{sec:metamodules-timeline}.
-
-\subsection{Motivating Example: Labeled Products and Regular Strings}\label{sec:metamodules-example}
-As a simple introductory example, let us consider \emph{labeled products} and \emph{regular strings}. In Verse, a simple labeled product type classifying conference papers can be defined like this:
-\begin{lstlisting}[numbers=none]
-let type Paper = lprodtype {SPCT
+Let us now turn our attention to the type structure of the Verse external language (EL).
+In contrast to other languages, which build in constructs like case types (shown briefly at the beginning of Sec.
+\ref{sec:examples}), tuples, records, objects and so on primitively, in Verse these can all be expressed as modes of use of \emph{metamodules} (which are distinct from, though conceptually related to, \emph{modules}).
+In fact, only polymorphic function types are built primitively into the Verse EL.
+
+This section is organized much like Section \ref{sec:syntax}.
+We begin with examples of type structure that we wish to express in Section \ref{sec:metamodules-example}, then discuss how existing approaches are insufficient in Section \ref{sec:metamodules-related}.
+Next, we show how metamodules (together with TSMs and TSLs) serve to address these problems in Section \ref{sec:metamodules-solution} and conclude with a timeline in Section \ref{sec:metamodules-timeline}.
+
+\subsection{Motivating Example: Labeled Tuples and Regular Strings}\label{sec:metamodules-example}
+As a simple introductory example, let us consider a type classifying conference papers:
+\begin{lstlisting}[numbers=none]
+let type Paper = ltuple {SPCT
    title : EPCTrstring /SURL.+EURL/
@@
 \end{lstlisting}
-The first row is labeled \lstinline{title} and specifies the regular string type \lstinline{rstring /SURL.*EURL/}.
-Regular string types classify values that behave as strings known to be in the regular language specified by the regular expression given as the type parameter.
-Here, we have that paper titles must be non-empty strings.
-The second row is labeled \lstinline{conf} and specifies a regular string type with two parenthesized groups, corresponding to the conference abbreviation and the year.
-The \lstinline{let type} construction defines a type synonym, \lstinline{Paper}. Alternatively, we could have used the following construction to declare \lstinline{Paper} \emph{generatively}:
-\begin{lstlisting}[numbers=none]
-lprodtype Paper {
-  SPCTtitle : EPCTrstring /SURL.+EURL/
-   SPCTconf : EPCTrstring /SURL([A-Z]+) (\d\d\d\d)EURL/
+The \lstinline{let type} construction defines a synonym, \lstinline{Paper}, for a type constructed by applying the \emph{type constructor} (or more concisely, \emph{tycon}) \lstinline{ltuple} to an ordered finite mapping from labels to types, delimited here by curly braces.
+Its first component is labeled \lstinline{title} and specifies the regular string type \lstinline{rstring /SURL.*EURL/}.
+Regular string types are constructed by applying the tycon \lstinline{rstring} to a regular expression pattern,  written literally here.
+They classify values that behave like strings in the corresponding regular language.
+So in this case, we have specified that paper titles must be non-empty strings.
+The second component of our labeled product type is labeled \lstinline{conf} and specifies a regular string type with two parenthesized groups, corresponding to the conference abbreviation and year.
+
+% Alternatively, we could have used the following construction to declare \lstinline{Paper} \emph{generatively}:
+% \begin{lstlisting}[numbers=none]
+% lprodtype Paper {
+%   SPCTtitle : EPCTrstring /SURL.+EURL/
+%    SPCTconf : EPCTrstring /SURL([A-Z]+) (\d\d\d\d)EURL/
+% }
+% \end{lstlisting}
+% The difference is in how type equality is handled. Two labeled product types with the same signature are equal, unless declared generatively, in which case type equality is nominal.
+
+We can introduce a value of type \lstinline{Paper} in an analytic position in one of two ways. We can omit the labels and provide component values positionally:
+\begin{lstlisting}[numbers=none]
+let exmpl : Paper = {"SUSAn Example PaperEUS"SCSS, ECSS"SUSEXMPL 2015EUS"}
+\end{lstlisting}
+Alternatively, we can include the component labels explicitly for readability or if we want to give the components in an alternative order:
+\begin{lstlisting}[numbers=none]
+let exmpl : Paper = {SCSSconf => ECSS"SUSEXMPL 2015EUS"SCSS, title => ECSS"SUSAn Example PaperEUS"SCSS}
+\end{lstlisting}
+%In either case, notice that we used literal syntax to introduce the component values, which are of regular string type.
+
+Given a value of type \lstinline{Paper}, we can project out a component value by providing  a  label:% like this:
+\begin{lstlisting}[numbers=none]
+let exmpl_conf = # <SNATconfENAT> exmpl 
+\end{lstlisting}
+Here, \lstinline{#} is an \emph{operator constructor} (or \emph{opcon}) parameterized by a static label, written literally here as \lstinline{<SNATconfENAT>}.
+The resulting \emph{operator}, \lstinline{# <SNATconfENAT>}, takes one \emph{argument}, \lstinline{exmpl}.
+
+We can also project captured groups out of a regular string using the \lstinline{#group} operator constructor, which is parameterized by a natural number referring to the group index:
+\begin{lstlisting}[numbers=none]
+let exmpl_conf_name = #group 0 exmpl_conf
+\end{lstlisting}
+Note that the variable \lstinline{exmpl_conf_name} has type \lstinline{rstring /SURL[A-Z]+EURL/}.
+
+Both labeled tuples and regular strings support a number of other basic operations.
+For example, labeled tuples can be concatenated (with any common components updated with the value on the right) using the \lstinline{ltuple+} operator:
+\begin{lstlisting}[numbers=none]
+let a : lprodtype {SPCTauthors : EPCTlist(rstring /SURL.+EURL/)} = {["SUSHarry Q. BovikEUS"]}
+let exmpl_paper_final = ltuple+ exmpl_paper a
+\end{lstlisting}
+We can drop a component using the operator constructor \lstinline{ltuple-}, which is parameterized by a static label:
+\begin{lstlisting}[numbers=none]
+let exmpl_paper_anon = ltuple- <SNATauthorsENAT> exmpl_paper_final
+\end{lstlisting}
+
+\subsection{Existing Approaches}\label{sec:metamodules-related}
+\subsubsection{Labeled Product Types}
+Recall that Verse builds in only nullary and binary products into its internal language, so the type structure described above for labeled tuples  is not directly expressible using IL primitives.
+The EL does not build in even these.
+However, so as to separate concerns, let us assume for the moment that regular string types are available in the EL: %So to consider existing approaches, we have to at least somewhat weaken our insistence on minimalism.
+\begin{lstlisting}[numbers=none]
+ (* type synonyms for concision *)
+ let type title_t = rstring /SURL.+EURL/
+ let type conf_t = rstring /SURL([A-Z]+) (\d\d\d\d)EURL/
+\end{lstlisting}
+
+\paragraph{Modules}
+One approach we might take is to attempt to express the type structure of any particular labeled tuple type using the module system.
+%For simplicity below, let us slightly weaken our insistence on minimalism and build binary products into the EL (and, for now, assume that regular string types are also built in in some suitable manner).%This is not strictly necessary -- we could use a Church-style encoding -- but this simplifies our examples.%then we could try to use the module system to express the type structure of any particular labeled tuple type.
+For our example above:
+For example, we might define the following signature for our example above:
+\begin{lstlisting}
+signature PAPER
+  type t
+  (* introduction without labels *)
+  val intro : title_t -> conf_t -> t
+  (* introduction with explicit labels, in both orders *)
+  val intro_title_conf : title_t -> conf_t -> t
+  val intro_conf_title : conf_t -> title_t -> t
+  (* projection *)
+  val prj_title : t -> title_t
+  val prj_conf : t -> conf_t
+\end{lstlisting}
+We've simply taken all possible valid introductory and projection operators and expressed them as functions, moving label parameters into identifiers.
+Even with the minimal EL we are assuming, we could  implement this signature using a Church-style encoding.
+The details are straightforward and omitted.
+Alternatively, if we only slightly relax slightly our insistence on minimalism and add binary products to the EL, we could use them as well:
+\begin{lstlisting}
+ module Paper : PAPER
+   type t = title_t * conf_t
+   fun intro_tpl title conf = (title, conf)
+   fun intro_title_conf title conf = (title, conf)
+   fun intro_conf_title conf title = (title, conf)
+   fun prj_title (title, conf) = title 
+   fun prj_conf (title, conf) = conf
+
+ type Paper = Paper.t
+ \end{lstlisting}
+%Compared to the four line type definition at the beginning of Sec.
+safe to say that this is an  unacceptable departure from the four line type definition at the beginning of Sec.
+\ref{sec:metamodules-example}.
+
+One problem is that this requires a volume of boilerplate code quadratic in the number of components of our labeled tuple type.
+It is also not entirely satisfying syntactically from the client's perspective.
+To avoid these problems and recover the syntax used previously, we could define a TSM for every such signature, though this too involves much boilerplate.
+
+More fundamentally, this encoding is limited to expressing the basic introductory and projection operators.
+We did not express all possible valid operators that arise from the opcon \lstinline{ltuple-} as functions, because to do so, we would need to construct an encoding of every labeled tuple type that might arise from dropping one or more components.
+This would require an exponential volume of boilerplate code.
+Finally, there is no way to define \lstinline{ltuple+} as a function or a finite collection of functions, because there are infinitely many extensions of any particular labeled product type.% of this operator would need to be implemented by clients manually, defeating the purpose of the operator.
+%For example, there are an arbitrary
+
+% \begin{lstlisting}[numbers=none]
+% signature PAPER
+%   type t
+%   val intro_tpl : rstring /.../ * rstring /.../ -> t
+%   val intro_conf_title : rstring /.../ * rstring /.../ -> t
+%   val intro_title_conf : rstring /.../ * rstring /.../ -> t
+%   val prj_conf : t -> rstring /.../
+%   val prj_title : t -> rstring /.../
+%  \end{lstlisting}
+
+\paragraph{Records and Tuples}
+If, instead of settling for such a module-based encoding, we further weaken our insistence on minimalism and primitively build in record/tuple types, as SML and many other languages have done, let us consider what might be possible.
+We note at the outset that records and tuples too are library constructs in Verse, but because they are so commonly built in to other languages, this situation is worth considering.%we can retreat from trying to use the module system, but have only slightly better options.
+
+The simplest approach we might consider taking is to directly map each labeled tuple type to a record type defining the same component mapping:
+\begin{lstlisting}[numbers=none]
+let type Paper_rcd = record {
+  title : title_t
+   conf : conf_t 
+}\end{lstlisting}
+Unlike labeled tuple types, record types are identified only up to component reordering.
+In other words, this mapping does not preserve certain type disequalities between labeled tuple types.
+Lacking positional information, it is not possible to omit component labels and rely on positional information when introducing a value of this type.%, because this information is not maintained by the type system.
+
+To work around this limitation if we do not otherwise care about preserving these type disequalities, we would need to define two conversion functions involving tuples:
+\begin{lstlisting}[numbers=none]
+fun Paper_of_tpl (title : title_t, conf : conf_t)
+  {title => title, conf => conf}
+fun tpl_of_Paper {title => title, conf => conf} = (title, conf)
+\end{lstlisting}
+For providers, this has syntactic cost linear in the number of components.
+For clients, we must explicitly invoke these functions, which incurs both a syntactic and run-time cost.
+We could define a TSM for every such type to reduce this cost to clients, though this is an additional cost to providers.%This also means that we cannot omit the labels and rely on positional information when introducing a value of a record type in an analytic position, as we did above with type \lstinline{Paper}.
+%If we instead use an unlabeled product type (which in ML is actually a record type with numeric labels, which are omitted syntactically), we lose the ability to use labels in the elimination form.
+
+If on the other hand we do care about preserving the type disequality mentioned, the workaround is even less elegant: we first need to define a record type with component labels tagged in some sufficiently unambiguous way by position, e.g.:
+\begin{lstlisting}[numbers=none]
+let type Paper = record {
+	__0_title : title_t
+	 __1_conf : conf_t 
+}\end{lstlisting}
+To avoid exposing this directly to clients, we need two auxiliary type definitions:
+\begin{lstlisting}[numbers=none]
+let type Paper_tpl = (title_t * conf_t)
+let type Paper_rcd = (* as above *)
+\end{lstlisting}
+and four conversion functions:
+\begin{lstlisting}[numbers=none]
+fun Paper_of_tpl (title : title_t, conf : conf_t) = {
+  __0_title => title,
+  __1_conf => conf
 }
-\end{lstlisting}
-The difference is in how type equality is handled. Two labeled product types with the same signature are equal, unless declared generatively, in which case type equality is nominal.
-
-In either case, we can introduce a value of type \lstinline{Paper} in an analytic position in one of two ways. We can omit the labels:
-\begin{lstlisting}[numbers=none]
-let exmpl : Paper = {"SUSAn Example PaperEUS"SCSS, ECSS"SUSEXMPL 2015EUS"}
-\end{lstlisting}
-Alternatively, we can include them explicitly for clarity:
-\begin{lstlisting}[numbers=none]
-let exmpl : Paper = {SCSStitle => ECSS"SUSAn Example PaperEUS"SCSS, conf => ECSS"SUSEXMPL 2015EUS"}
-\end{lstlisting}
-
-We can then project a row out of such a value using the row projection operator:
-\begin{lstlisting}[numbers=none]
-let exmpl_conf = # `SNATconfENAT exmpl 
-\end{lstlisting}
-Here, \lstinline{#} is an operator parameterized by a static label, written literally \lstinline{`SNATconfENAT}, and taking one argument, \lstinline{exmpl}.
-
-We can also project captured groups out of a regular string using the \lstinline{#group} operator, which is parameterized by a group index:
-\begin{lstlisting}[numbers=none]
-let exmpl_conf_name = #group 0 exmpl_conf
-\end{lstlisting}
-
-The variable \lstinline{exmpl_conf_name} has type \lstinline{rstring /SURL[A-Z]+EURL/}.
-
-Both labeled products and regular strings support a number of other operations. For example, labeled products can be concatenated (with any common rows updated with the value on the right):
-\begin{lstlisting}[numbers=none]
-let a : lprodtype {SPCTauthors : EPCTlist(rstring /SURL.+EURL/)} = {["SUSHarry Q. BovikEUS"]}
-let exmpl_paper_final = lprod+ exmpl_paper a
-\end{lstlisting}
-We can drop a row as well:
-\begin{lstlisting}[numbers=none]
-let exmpl_paper_anon = lprod- exmpl_paper_final `SNATauthorsENAT
-\end{lstlisting}
-
-\subsection{Existing Approaches}\label{sec:metamodules-related}
-\subsubsection{Records and Tuples}
-Recall that Verse builds in only nullary and binary products into its internal language, so the type structure described above for labeled products  is not directly expressible using IL primitives.
-If we weaken slightly our insistence on minimalism and built in record/tuple types, as in other functional languages, we have  options that get close to (but do not  exactly achieve) this type structure.
-In particular, unlike labeled product types, record types are identified up to row reordering.
-Using a record with the same row labels would prevent us from omitting the labels and relying on positional information when introducing a value of record type in an analytic position, as we did with type \lstinline{Paper} above.
-%If we instead use an unlabeled product type (which in ML is actually a record type with numeric labels, which are omitted syntactically), we lose the ability to use labels in the elimination form.
-
-One workaround is to replace each labeled product type in the program with a record type with row labels tagged by position (recovering the type disequality we seek to express with labeled product types) and conversion functions from the corresponding unlabeled product type and untagged record type to this type.
-This partially recovers some of the syntactic cost we seek, but introduces much additional  boilerplate as well as run-time cost.
-We are interested in such issues, so this is not an entirely satisfying solution.
+fun tpl_of_Paper {__0_title => title, __1_conf => conf} = (title, conf)
+fun Paper_of_rcd {title => title, conf => conf} = {
+  __0_title => title,
+  __1_conf => conf
+}
+fun rcd_of_Paper {__0_title => title, __1_conf => conf} = {
+  title => title,
+  conf => conf
+}
+\end{lstlisting}
+Clients again bear the cost of applying these conversion functions ahead of every operation and providers again bear the burden of defining this boilerplate (which is again linear in the number of components, albeit with a constant factor that is again not easy to dismiss).
+%We are interested in all such pragmatic issues, so this is not an entirely satisfying solution.
 % In any case, records, tuples and labeled tuples are all library constructs in Verse.
-Moreover, this still does not give us support for the extra operations, like concatenation,  shown above.
-
-\subsubsection{Refinement Types}
-For regular strings, one could instead use standard strings (which themselves may be lists or vectors of characters together with derived syntax), inserting run-time checks around introductory forms and into other operations (with one exception, discussed below).
-This would, of course,  weaken the static semantics and add both syntactic and run-time cost.
-To recover the static guarantees of regular strings, we could treat membership of a string in a specified regular language as a static verification condition, moving the logic governing regular string types out of the language and into an external \emph{refinement type system} \cite{Freeman91}.
-
-One problem with this approach is that there is no direct analog to the group projection operator described above.
-The static and dynamic semantics of this operator depend on information available only in the static type of the regular string (i.e.
-the position of the captured group).
-Refinement types are strictly a layer above the language, and thus cannot affect its dynamic semantics.
-
-A related problem is that there is no feasible way to use refinement types to introduce type disequalities that can later affect the internal representation of a value.
-All standard strings are represented uniformly.
-The fact that some value of type \lstinline{string} can be given a refined type \lstinline{rstring /A|B/} cannot be justification to locally alter its representation (e.g.
-to a single bit in this case).
-For regular strings, there may be limited benefit to doing this, but in general, this can have significant performance ramifications.
-For example, it is common to justify the omission of a primitive type of ``unsigned integers representable in 32 bits'' by the fact that this is ``merely'' a refinement of arbitrary precision integers, but in so doing one loses the ability to use a machine representation more suitable to this invariant.
-%Refinement types cannot be used for this purpose without the possibility of conflict.
-This is also the reason why one cannot viably do away with primitive record types by claiming that they are refinements of finite mappings.
-
-Put another way, refinement types are useful when one only needs to make finer static distinctions within an existing type.
+
+As with our modular encoding above, neither of these workarounds provides us with any way to uniformly express the more interesting operations, like  concatenation, that we discussed earlier.
+
+\subsubsection{Regular Strings}
+Let us now turn to regular string types.
+
+\paragraph{Run-Time Checks}
+The most common alternative to regular strings in existing languages is to instead use standard strings (which themselves may be expressed as lists or vectors of characters together with derived syntax), inserting run-time checks around operations to maintain the regular string invariant dynamically.
+This clearly does not express the static semantics of regular strings, and incurs syntactic and run-time cost.
+
+\paragraph{Type Refinements}
+We might instead try to express the static semantics of regular strings by using standard strings and moving the logic governing regular string types into an external system of \emph{type refinements}  \cite{Freeman91}.
+%This treats membership of a string in a specified regular language exclusively as a static verification condition.
+These systems typically examine refinement specifications supplied in comments:
+\begin{lstlisting}[numbers=none]
+let conf : string (* ~ rstring /([A-Z]+) (\d\d\d\d)/ *) = "EXMPL 2015"
+\end{lstlisting}
+One problem with this is that there is no way to express the group projection operator described above. We cannot define a function \lstinline{#group} that can be used like this:
+\begin{lstlisting}[numbers=none]
+let conf_venue = #group 0 conf
+\end{lstlisting}
+The reason is that this function would need information that is  available only in the type refinement of each regular string, i.e.
+the positions of the captured groups.
+%Refinement types are strictly a layer above the language, and thus cannot affect its dynamic semantics.
+
+A closely related problem is that type refinements do not introduce type disequalities.
+As a result, compilers cannot use the type refinements to optimize the representation of a value.
+For example, the fact that some value of type \lstinline{string} can be given the refined type \lstinline{rstring /A|B/} cannot be justification to locally alter its representation to, for example, a single bit, both because it could later be passed into a function expecting a standard string and because there are many other possible refinements.
+For regular strings, there may be limited benefit to controlling representation differentially, but in other situations, this can have significant performance ramifications.
+For example, it is quite common to justify the omission of a primitive type of ``non-negative integers representable in 32 bits'' by the fact that this is ``merely'' a refinement of integers, but in so doing one loses the ability to use a machine representation more suitable to this particular static invariant.%Refinement types cannot be used for this purpose without the possibility of conflict.
+%This is also the reason why one cannot viably do away with primitive record types by claiming that they are refinements of finite mappings with labels as keys.
+
+%In summary, type refinements are appropriate when one needs only to make finer static distinctions within an existing type.
 When one needs to introduce new operators or type disequalities that have ``downstream'' impact, refinement types are not the right solution.%Refinement types are useful when no non-trivial operators need to be introduced into the language and the refinement has no impact on representation, but the type system itself must be extended in some way in more advanced situations.
 
+\paragraph{Modules}
+To create type disequalities, we might consider again turning to the module system, defining a signature for each regular string type that we wish to work with.
+Again, this requires a non-trivial amount of boilerplate code:
+\begin{lstlisting}[numbers=none]
+signature TITLE
+  type t
+  val intro : string -> t
+  val strcase : t -> (unit -> 'a) -> (string * string -> 'a) -> 'a
+
+signature CONF
+  type t
+  val intro : string -> t
+  val strcase : t -> (unit -> 'a) -> (string * string -> 'a) -> 'a
+  val prj_group_0 : t -> string
+  val prj_group_1 : t -> string
+\end{lstlisting}
+
+To ensure that applications of the introductory operators are correct statically, we would need to couple this with a system of refinement types.
+
+Unfortunately, this again reveals itself to be too limited a solution when we need other operators, e.g.
+concatenation.
+Expressing concatenation uniformly using a single function or a finite set of functions is impossible, for the same reason that applied to \lstinline{ltuple+} above, i.e.
+we would need to define one such function for every pair of possible regular string types (of which there are infinitely many).
+
 \subsection{Contributions}\label{sec:metamodules-approach}
 
-An important point is that labeled products, as well as other ``record-like'' constructs, can  be expressed by a \emph{type-directed translation} targeting our IL (and indeed, many compilers take advantage of facts like this to simplify the number of cases that must be considered downstream).
+%An important point is that labeled products, as well as other ``record-like'' constructs, can  be expressed by a \emph{type-directed translation} targeting our IL (and indeed, many compilers take advantage of facts like this to simplify the number of cases that must be considered downstream).
 %it is inaccurate to state that they are merely syntactic sugar over such a language.
@@
 %Other dialects are justified by a desire to provide more special-purpose semantic constructs. For example, 
-Substantially more sophisticated examples require at most only a very slightly more capable intermediate language (the IL of popular virtual machines, even those with flaws like ubiquitous \texttt{null} references, e.g.
+%Substantially more sophisticated examples require at most only a very slightly more capable intermediate language (the IL of popular virtual machines, even those with flaws like ubiquitous \texttt{null} references, e.g.
 JVM or CIL bytecode, are also sufficient if used strictly as translation targets).
 
-Metamodules introduce type and operator constructors parameterized by arbitrary \emph{static values} and metaprogrammatically express the logic governing their typechecking and  translation to the IL using the Verse \emph{static language} (SL).
+Metamodules introduce type and operator constructors parameterized by arbitrary \emph{static values} and metaprogrammatically express the logic governing their typechecking and  translation to the IL.
 Metamodules are classified by \emph{metasignatures}.
@@
 \begin{lstlisting}
-metasignature LPROD
-  tycon c of LblTyList
+metasignature LTUPLE
+  tycon c of LblTyMap
   ana opcon intro_unlabeled
@@
   syn opcon # of Lbl
-\end{lstlisting}\todo{reworking active tycons as metasignatures and metastructures is compelling, but I need to figure it out. will continue writing.}
-
-% The semantics uses internal type abstraction to guarantee that metamodule providers can reason modularly about translation invariants, just as it uses external type abstraction to guarantee that module providers can reason modularly about representation invariants.%For example, a metamodule that expresses the type structure of records would introduce a type constructor  \texttt{record} parameterized by a static mapping from labels to types, and two operators: the introductory operator \texttt{intro}, parameterized by a list of row labels, and the row projection operator \texttt{prj}, parameterized by a row label.
+syntax ltpl(L ~ LTUPLE, idx : LblTyList) at L.c(idx)
+  ...
+metamodule LTuple ~ LTUPLE with syntax ltpl for c
+  tycon c(idx : LblTyList) ~> ...
+  ana opcon intro_unlabeled(ty : Ty, args : List(Arg)) ~> ...
+  ana opcon intro_labeled(lbls : List(Lbl), ty : Ty, args : List(Arg)) ~> ...
+  syn opcon #(lbl, args) ~> ...
+let tycon ltuple = LTuple.c
+let opcon # = LTuple.#
+\end{lstlisting}
+
+c is a tycon parameterized by static maps from labels to types.
+
+intro unlabeled is a trivially parameterized opcon, i.e. just one operator.
+
+intro labeled is an opcon parameterized by a list of labels.
+
+\# is an opcon parameterized by a single label.
+
+ltpl is a TSM parameterized by a metamodule and a static value
+
+LTuple uses ltpl as a TSL
+
+\begin{lstlisting}
+\end{lstlisting}
+% The semantics uses internal type abstraction to guarantee that metamodule providers can reason modularly about translation invariants, just as it uses external type abstraction to guarantee that module providers can reason modularly about representation invariants.%For example, a metamodule that expresses the type structure of records wouldLblTyList introduce a type constructor  \texttt{record} parameterized by a static mapping from labels to types, and two operators: the introductory operator \texttt{intro}, parameterized by a list of row labels, and the row projection operator \texttt{prj}, parameterized by a row label.
 It might then choose to internally represent records by binary products, which the IL provides.

### 2015-04-09 thesis-proposal/problem.tex
rw

@@
 Instead, its users are able to modularly  express a wide variety of useful constructs in libraries, using the comparatively small set of primitives that the language builds in.
-Functional programming languages like Standard ML (SML) and Haskell descend directly from elegant typed lambda calculi, diverging mainly in that they strive to present a more pragmatic concrete syntax and type structure to programmers.
+Functional programming languages like Standard ML (SML) and Haskell descend directly from mathematically elegant typed lambda calculi, diverging mainly in that they present a more pragmatic concrete syntax and type structure to programmers.
 In other words, their design considers important human factors in addition to fundamental metatheoretic issues.
-For example, both languages build in record types, generalizing the nullary and binary product types more common in simpler calculi, because labels are cognitively useful.
+For example, they build in record types, generalizing the nullary and binary product types more common in simpler calculi, because labels are cognitively useful.
+Similarly, they build in derived concrete syntax for the list datatype.
 %Minimal typed lambda calculi play a central role in the study of programming languages, and serve as direct inspiration for  functional programming languages like ML.
@@
 In fact, tools for constructing new ``domain-specific'' dialects seem only to be increasingly popular. 
-{This calls for an investigation}: why is it that well-informed programmers and researchers are so often unsatisfied with expressing the constructs that they need in libraries, as modes of use of the ``general-purpose'' functional primitives available today?
+{This calls for an investigation}: why is it that well-informed programmers and researchers are so often unsatisfied with expressing the constructs that they need in libraries, as modes of use of the ``general-purpose'' functional primitives already available today?
 %Let us investigate why this is.
@@
 For example,  we will consider regular expression patterns expressed using abstract data types (in ML, a mode of use of the module system) in Section \ref{sec:syntax}.
-This  will give  us precisely the semantics that we need, but syntactically, it will leave something to be desired.
-To avoid this syntactic cost, library providers are frequently tempted to construct a \emph{derived syntactic dialect}, i.e.
-a dialect specified by context-free elaboration to the existing language, that introduces derived syntax specific to their library.
+This  will give  us precisely the semantics that we seek to express, but syntactically, the usual approach of using strings to introduce patterns actually leaves much to be desired.
+In response, we may be tempted to construct a \emph{derived syntactic dialect}, i.e.
+a dialect specified by context-free elaboration to the existing language, that introduces derived syntax specific to our library, just as languages like SML have introduced derived syntax for constructs in the standard library, like lists.
+This is not at all an isolated example.
 For example, Ur/Web is a  derived syntactic dialect of Ur (itself descended from ML) that builds in derived syntax for SQL queries, HTML elements and other datatypes common in web programming \cite{conf/popl/Chlipala15}.
-In fact, nearly all major programming languages primitively build in derived syntax for constructs that can otherwise be expressed in the standard library, e.g.
+%In fact, nearly all major programming languages primitively build in derived syntax for constructs that can otherwise be expressed in the standard library, e.g.
 derived list syntax in SML \cite{harper1997programming,mthm97-for-dart}.
-Tools like Camlp4 \cite{ocaml-manual} and Sugar* \cite{erdweg2011sugarj,erdweg2013framework} (which we will return to later) help lower the engineering costs of constructing such dialects, contributing to their proliferation.% contributing to their proliferation.
+Tools like Camlp4 \cite{ocaml-manual} and Sugar* \cite{erdweg2011sugarj,erdweg2013framework} (which we will return to later) are designed to lower the engineering costs of constructing such dialects, contributing to their proliferation.% contributing to their proliferation.
 
@@
 Indeed, perusal of the proceedings of any major programming languages conference will typically yield several more examples.
-Tools like proof assistants and logical frameworks support specifying and reasoning metatheoretically about dialects like these, and tools like compiler generators and language workbenches lower the implementation cost, again contributing to their increasing proliferation.
+Tools like proof assistants and logical frameworks support specifying and reasoning metatheoretically about dialects like these, and tools like compiler generators and language workbenches lower their implementation cost, again contributing to their continued proliferation.
 
@@
 given two dialects that are specified by judgements of arbitrary form, it is not clear what it would even mean to combine them.
-Even when the dialects are specified using a formalism that does provide some way to na\"ively combine two dialects, there is generally no guarantee that the composition will conserve important syntactic and semantic properties.
+Even when the dialects are specified using some formalism that does provide a means to na\"ively combine two dialects, there is generally no guarantee that the composition will conserve important syntactic and semantic properties.
 %There is no well-defined mechanism for constructing such a ``combined language'' in general.
@@
 
-Dialects can have an indirect impact on large-scale software development because the language designers that control comparatively popular languages, like Ocaml and Scala, can be convinced to incorporate  ideas from dialects  into backwards compatible language revisions.
+Dialects do have an indirect impact on large-scale software development because the language designers that control comparatively popular languages, like Ocaml and Scala, can be convinced to incorporate  ideas from dialects  into backwards compatible language revisions.
 %These decisions are increasingly influenced by community processes, e.g.
@@
 %Dialects thus serve the role of rhetorical vehicles for new ideas, rather than direct artifacts. 
-These languages have, in this way, become more expressive over time.
-However, this expressiveness has come at a cost: they have also ballooned in size.
-Yet because there appears to be a ``long tail'' of potentially useful primitives (for derived syntax, our group has  gathered initial data speaking to this \cite{TSLs}),  primitives that are only situationally useful, or that trigger aesthetic disagreements between different factions of the community, are still often left languishing in ``toy'' dialects.
+In this this way, these languages have become more general over time.
+However, this  has come at a cost: they have also ballooned in size.
+And yet, because there appears to be a ``long tail'' of potentially useful primitives (for derived syntax, our group has  gathered initial data speaking to this \cite{TSLs}),  primitives that are only situationally useful, or that trigger aesthetic disagreements between different factions of the community, are still often left languishing in ``toy'' dialects.
 Recalling the words of  Reynolds, which are as relevant today as they were nearly half a century ago \cite{Reynolds70}:%Because there is no data about how useful a construct is in practice until it is included in a language like this, decisions about which constructs to include are often informed by little more than intuition.
@@
 The other path forward is to continue on in the search for even more general language primitives, i.e.
-primitives that are so expressive that they permit us to degrade a broad class of existing primitives, including those found in dialects, to modularly composable library constructs, where they can be evaluated on their individual merits by programmers and used together without the possibility of conflict.
+primitives that are so expressive that they permit us to degrade a broad class of existing primitives, including those found in dialects, to modularly composable library constructs, so that they can be evaluated on their individual merits by programmers and used together without the possibility of conflict.
 % This should render the construction of new dialects increasingly unnecessary.
-Encouragingly, such new primitives do occasionally arise.
-For example, Ocaml recently added support for  ``generalized algebraic data types'' (GADTs), based on research on guarded recursive datatype constructors \cite{XiCheChe03}.
+Encouragingly, primitives that have this effect on a language do occasionally arise.
+For example, Ocaml in a recent revision added support for  ``generalized algebraic data types'' (GADTs), based on research on guarded recursive datatype constructors \cite{XiCheChe03}.
 Using GADTs, Ocaml was able to move some of the machinery for typechecking operations that use format strings, like \texttt{sprintf}, out of the language and into the standard library (unfortunately, some machinery remains built in).
-
 Our broad aim is to introduce a new functional programming language called Verse\footnote{We distinguish Verse from Wyvern, which is the language referred to in prior publications about some of the work being proposed here, because Wyvern is a group effort evolving independently in some important ways.} that takes even more dramatic steps down this second path.
@@
 list syntax as built in to SML, by giving library providers static control over the parsing and elaboration of delimited segments of concrete syntax.
-\item \textbf{Metamodules}, which we introduce in Section \ref{sec:metamodules}, give library providers static hooks directly into the semantics of the Verse external language, allowing  them to define, for example, the type structure of records (and various ``record-like'' constructs, as discussed above) by type-directed translation to the more minimal internal language (which builds in nullary and binary products, but not records).
+\item \textbf{Metamodules}, which we introduce in Section \ref{sec:metamodules}, give library providers static hooks directly into the type-directed translation protocol that governs the Verse external language.
+For example, library providers will be able to define a metamodule that introduces extensible labeled product types and their associated operators, defining their semantics by type-directed translation to the internal language, which builds in only nullary and binary product types.
 \end{itemize} 
@@
 The key challenge in the design of these primitives will come in ensuring that they are metatheoretically well-behaved, given that they aim to give library providers decentralized control over aspects of the language that have typically been under the monolithic control of the language designer.
-If we are not careful, many of the problems discussed above as inherent to combining distinct language dialects could simply manifest themselves in the semantics of these primitives.\footnote{This is why languages  like Verse are often called \emph{extensible languages}, though this is somewhat of a misnomer.
+If we are not careful, many of the problems discussed earlier as inherent to combining distinct language dialects could simply manifest themselves in the semantics of these primitives.\footnote{This is why languages  like Verse are often called \emph{extensible languages}, though this is somewhat of a misnomer.
 The chief characteristic of an extensible language is that it \emph{doesn't} need to be extended in situations where other languages would need to be extended.
 We will avoid this somewhat confusing terminology.} Our main technical contributions will come in showing how to address these problems in a principled manner.
-In particular, syntactic conflicts will be impossible by construction and  we will validate the code  statically generated by TSMs and metamodules to maintain a hygienic type discipline and, most uniquely, powerful modular reasoning principles -- library providers will be able to reason about the constructs that they have defined in isolation, and library clients will be able to use them safely in any combination, without the possibility of conflict.\footnote{We  assume that simple naming conflicts can be avoided extrinsically, e.g.
+In particular, syntactic conflicts will be impossible by construction and  we will validate the code  statically generated by TSMs and metamodules to maintain a hygienic type discipline and, most uniquely, powerful modular reasoning principles.
+In other words, library providers will have the ability to reason about the constructs that they have defined in isolation, and library clients will be able to use them safely in any combination, without the possibility of conflict.\footnote{We  assume that simple naming conflicts can be avoided extrinsically, e.g.
 by using a URI-based naming scheme as in the Java ecosystem.}
@@
 The module language is based directly on SML's module language,  which we assume working familiarity with for the purposes of this proposal \cite{harper1997programming,MacQueen:1984:MSM:800055.802036}.
-We will give examples of its use in both Section \ref{sec:syntax} and Section \ref{sec:metamodules}, but do not plan to specify it in detail.
+We will give examples of its use in both Section \ref{sec:syntax} and Section \ref{sec:metamodules}, but do not plan to specify it in detail in the thesis.\todo{reminder that we are not going to re-specify the module system}
 

### 2015-04-12 thesis-proposal/problem.tex
updated sec5

@@
 Instead, its users are able to modularly  express a wide variety of useful constructs in libraries, using the comparatively small set of primitives that the language builds in.
-Functional programming languages like Standard ML (SML) and Haskell descend directly from mathematically elegant typed lambda calculi, diverging mainly in that they present a more pragmatic concrete syntax and type structure to programmers.
-In other words, their design considers important human factors in addition to fundamental metatheoretic issues.
-For example, they build in record types, generalizing the nullary and binary product types more common in simpler calculi, because labels are cognitively useful.
-Similarly, they build in derived concrete syntax for the list datatype.
+Functional programming languages like Standard ML (SML) and Haskell descend directly from  elegant typed lambda calculi, diverging mainly in that they consider  various human factors in addition to fundamental metatheoretic issues in their design.
+For example, they both build in record types, generalizing the nullary and binary product types that are more common in simpler calculi, because labeled components are cognitively useful.
+Similarly, they both build in derived  syntax for the standard list datatype.% concision and readability.
 %Minimal typed lambda calculi play a central role in the study of programming languages, and serve as direct inspiration for  functional programming languages like ML.
@@
 
-Ideally, transitioning from a minimal calculus to a pragmatic programming language would require introducing only a limited number of primitive ``embellishments'' like this.
-The community around the language would then be able to redirect its efforts exclusively toward the development of a wide variety of useful libraries.
-Alas, a language design that achieves this ideal has proven elusive, as evidenced by the diverse array of {dialects} that continue to proliferate around all major languages, functional languages included.
+Ideally, the transition from  a minimal calculus to a pragmatic ``full-scale'' programming language would require only a limited number of primitive ``embellishments'' like these.
+The community around the language could then direct its efforts almost exclusively toward the development of useful libraries.
+Unfortunately, a single language design that achieves this ideal has proven elusive, as evidenced by the diverse array of {dialects} that continue to proliferate around all major contemporary languages, functional languages included.
 % For example, let us consider the many dialects of ML.
@@
 %Tools for  constructing new ``domain-specific'' dialects also continue to proliferate. % The situation is similar across all major language lineages. 
-In fact, tools for constructing new ``domain-specific'' dialects seem only to be increasingly popular. 
-{This calls for an investigation}: why is it that well-informed programmers and researchers are so often unsatisfied with expressing the constructs that they need in libraries, as modes of use of the ``general-purpose'' functional primitives already available today?
+In fact, tools that make constructing ``domain-specific'' language dialects easier appear to be increasingly popular. 
+{This calls for an investigation}: why are well-informed programmers and researchers so often unable to satisfyingly express the constructs that they need in libraries, as modes of use of the ``general-purpose'' functional primitives already available in major contemporary languages today?
 %Let us investigate why this is.
@@
 %Syntactic cost is often assessed qualitatively \cite{green1996usability}, though quantitative metrics can be defined.
-For example,  we will consider regular expression patterns expressed using abstract data types (in ML, a mode of use of the module system) in Section \ref{sec:syntax}.
-This  will give  us precisely the semantics that we seek to express, but syntactically, the usual approach of using strings to introduce patterns actually leaves much to be desired.
-In response, we may be tempted to construct a \emph{derived syntactic dialect}, i.e.
-a dialect specified by context-free elaboration to the existing language, that introduces derived syntax specific to our library, just as languages like SML have introduced derived syntax for constructs in the standard library, like lists.
-This is not at all an isolated example.
+For example,  we consider regular expression patterns expressed using abstract data types (in ML, a mode of use of the module system) in Sec.
+\ref{sec:syntax}.
+This  gives  us precisely the semantics that we seek to express, but syntactically, we argue that the usual approach of using strings to introduce patterns leaves much to be desired.
+To avoid this, we are tempted to introduce derived syntax specific to our library by constructing a \emph{derived syntactic dialect}, i.e.
+a dialect specified by context-free elaboration to the existing language.
+This is far from an isolated example.
 For example, Ur/Web is a  derived syntactic dialect of Ur (itself descended from ML) that builds in derived syntax for SQL queries, HTML elements and other datatypes common in web programming \cite{conf/popl/Chlipala15}.
@@
 derived list syntax in SML \cite{harper1997programming,mthm97-for-dart}.
-Tools like Camlp4 \cite{ocaml-manual} and Sugar* \cite{erdweg2011sugarj,erdweg2013framework} (which we will return to later) are designed to lower the engineering costs of constructing such dialects, contributing to their proliferation.% contributing to their proliferation.
+Tools like Camlp4 \cite{ocaml-manual} and Sugar* \cite{erdweg2011sugarj,erdweg2013framework} (which we will return to later) are designed to lower the engineering costs of constructing derived syntactic dialects, contributing to their proliferation.% contributing to their proliferation.
 
@@
 
-More advanced dialects introduce  new type structure, going beyond what is possible by context-free elaboration to the existing language.
-For example, as  mentioned above, Standard ML goes beyond the nullary and binary product types found in simpler calculi, instead exposing record types.
+More advanced dialects introduce  new type structure, going beyond what is possible by context-free elaboration to an existing language.
+For example, as mentioned above, SML goes beyond nullary and binary product types to expose record types.
+%Note however that the dynamic semantics of record types can, of course, be expressed by translation to a minimal calculus.
 Other dialects have explored  ``record-like'' primitives that support functional update and extension operators, width and depth coercions (sometimes implicit)%\cite{Cardelli:1984:SMI:1096.1098}
-, mutable rows, methods, prototypic dispatch and various other embellishments.
-Ocaml builds in a  sophisticated object system, polymorphic variants and  conveniences like logic for typechecking operations that use format strings, e.g.
+, mutable components, methods, prototypic dispatch and various other embellishments.
+Ocaml builds in an object system, polymorphic variants and  conveniences like logic for typechecking operations that use format strings, e.g.
 $\mathtt{sprintf}$ \cite{ocaml-manual}.
@@
 Indeed, perusal of the proceedings of any major programming languages conference will typically yield several more examples.
-Tools like proof assistants and logical frameworks support specifying and reasoning metatheoretically about dialects like these, and tools like compiler generators and language workbenches lower their implementation cost, again contributing to their continued proliferation.
+Tools like proof assistants and logical frameworks support specifying and reasoning metatheoretically about dialects like these, and tools like compiler generators and language workbenches lower the cost of implementing them, again contributing to their continued proliferation.
 
@@
 %Dialects thus serve the role of rhetorical vehicles for new ideas, rather than direct artifacts. 
-In this this way, these languages have become more general over time.
-However, this  has come at a cost: they have also ballooned in size.
-And yet, because there appears to be a ``long tail'' of potentially useful primitives (for derived syntax, our group has  gathered initial data speaking to this \cite{TSLs}),  primitives that are only situationally useful, or that trigger aesthetic disagreements between different factions of the community, are still often left languishing in ``toy'' dialects.
+However, this  has come at a cost: these languages have ballooned in size.
+And yet, because there appears to be a ``long tail'' of potentially useful primitives (for derived syntax, our group has  gathered initial data speaking to this \cite{TSLs}),  primitives that are only situationally useful, or that trigger aesthetic disagreements between different factions of the community, are still quite often left languishing in ``toy'' dialects.
 Recalling the words of  Reynolds, which are as relevant today as they were nearly half a century ago \cite{Reynolds70}:%Because there is no data about how useful a construct is in practice until it is included in a language like this, decisions about which constructs to include are often informed by little more than intuition.
@@
 % This should render the construction of new dialects increasingly unnecessary.
-Encouragingly, primitives that have this effect on a language do occasionally arise.
+Encouragingly, primitives that have such a ``slimming'' effect on a language do occasionally arise.
 For example, Ocaml in a recent revision added support for  ``generalized algebraic data types'' (GADTs), based on research on guarded recursive datatype constructors \cite{XiCheChe03}.
@@
 
-Verse has a module system taken directly from SML, but its core language is organized in a novel manner around a minimal typed lambda calculus, called the Verse \emph{internal language}, discussed in Section \ref{sec:verse}.
-In lieu of typical {external} (i.e.
+Verse has a module system taken directly from SML, but its core language is specified by a \emph{type-directed translation semantics} targeting a minimal typed lambda calculus, called the Verse \emph{internal language}, introduced in Sec.
+\ref{sec:verse}.
+In lieu of typical \emph{external} (i.e.
 programmer-facing) derived syntax and  type structure, Verse introduces two novel primitive constructs:
 \begin{itemize}
-\item \textbf{Typed syntax macros} (TSMs), which we introduce in Section \ref{sec:syntax}, subsume the need for derived syntax specific to library constructs, e.g.
-list syntax as built in to SML, by giving library providers static control over the parsing and elaboration of delimited segments of concrete syntax.
-\item \textbf{Metamodules}, which we introduce in Section \ref{sec:metamodules}, give library providers static hooks directly into the type-directed translation protocol that governs the Verse external language.
-For example, library providers will be able to define a metamodule that introduces extensible labeled product types and their associated operators, defining their semantics by type-directed translation to the internal language, which builds in only nullary and binary product types.
+\item \textbf{Typed syntax macros} (TSMs), introduced in Sec.
+\ref{sec:syntax}, subsume the need to primitively build in derived concrete syntax specific to library constructs, e.g.
+list syntax as in SML or XML syntax as in Scala and Ur/Web, by giving library providers static control over the parsing and elaboration of delimited segments of concrete syntax.
+\item \textbf{Metamodules}, introduced in Sec.
+\ref{sec:metamodules}, subsume the need to primitively build in the type structure of constructs like records (and variants thereof) by giving library providers static hooks directly into the type-directed translation semantics governing the Verse external language.
+For example, we will discuss how a library provider can define a single metamodule that defines \emph{labeled tuple types} together with the operators that allow clients to introduce, project out of and concatenate values of these types.
+The internal language builds in only nullary and binary product types primitively.
 \end{itemize} 
-Both TSMs and metamodules can be understood as \emph{metaprogramming} primitives, because they involve static code generation.
+Both TSMs and metamodules can be understood as \emph{metaprogramming} primitives, because they involve programmatic static code generation.
 %We will  also introduce a simple variant of each of these primitives that leverages Verse's support for local type inference to further reduce syntactic cost in certain common situations.
 
-The key challenge in the design of these primitives will come in ensuring that they are metatheoretically well-behaved, given that they aim to give library providers decentralized control over aspects of the language that have typically been under the monolithic control of the language designer.
-If we are not careful, many of the problems discussed earlier as inherent to combining distinct language dialects could simply manifest themselves in the semantics of these primitives.\footnote{This is why languages  like Verse are often called \emph{extensible languages}, though this is somewhat of a misnomer.
+The key challenge in the design of these primitives will come in ensuring that they are metatheoretically well-behaved, given that they aim to give library providers decentralized control over aspects of Verse's concrete syntax and type structure that, in other languages, are under the monolithic control of the language designer.
+If we are not careful, many of the problems discussed earlier as inherent to combining distinct language dialects could simply shift into the semantics of these primitives.\footnote{This is why languages  like Verse are often called \emph{extensible languages}, though this is somewhat of a misnomer.
 The chief characteristic of an extensible language is that it \emph{doesn't} need to be extended in situations where other languages would need to be extended.
 We will avoid this somewhat confusing terminology.} Our main technical contributions will come in showing how to address these problems in a principled manner.
-In particular, syntactic conflicts will be impossible by construction and  we will validate the code  statically generated by TSMs and metamodules to maintain a hygienic type discipline and, most uniquely, powerful modular reasoning principles.
-In other words, library providers will have the ability to reason about the constructs that they have defined in isolation, and library clients will be able to use them safely in any combination, without the possibility of conflict.\footnote{We  assume that simple naming conflicts can be avoided extrinsically, e.g.
+In particular, syntactic conflicts will be made impossible by construction and  we will validate the code  statically generated by TSMs and metamodules to maintain a \emph{hygienic type discipline} and, most uniquely, powerful \emph{modular reasoning principles}.
+In other words, library providers will have the ability to reason about the constructs that they define in isolation, and clients will be able to use them safely in any combination, without the possibility of conflict.\footnote{We  assume that naming conflicts can be avoided extrinsically, e.g.
 by using a URI-based naming scheme as in the Java ecosystem.}
@@
 \section{Summary of Judgements}\label{sec:verse}
-To situate ourselves within a formal framework, let us begin with a brief summary of how Verse is organized and specified.
-Verse consists of a \emph{module language} atop a \emph{core language}.
-The module language is based directly on SML's module language,  which we assume working familiarity with for the purposes of this proposal \cite{harper1997programming,MacQueen:1984:MSM:800055.802036}.
-We will give examples of its use in both Section \ref{sec:syntax} and Section \ref{sec:metamodules}, but do not plan to specify it in detail in the thesis.\todo{reminder that we are not going to re-specify the module system}
-
-The core language is organized much like the first stage of a type-directed compiler (e.g.
-the TIL compiler for Standard ML \cite{tarditi+:til-OLD}), consisting of a user-facing \emph{typed external language} (EL) specified by type-directed translation to a minimal \emph{typed internal language} (IL).\footnote{Note that we will use the terms ``expression'' and ``type'' without qualification to refer to  external expressions and types, respectively.
-} The main judgements in the specification of the EL take the following form (omitting various contexts for now):
+To situate ourselves within a formal framework, let us begin with a brief summary of how Verse is organized and specified. Verse consists of a \emph{module language} atop a \emph{core language}.
+
+The Verse module language is based directly on SML's module language, with which we assume a working familiarity for the purposes of this proposal \cite{harper1997programming,MacQueen:1984:MSM:800055.802036} (a related module language, e.g.
+Ocaml's, would work just as well for our purposes).
+We will give examples of its use in Sections \ref{sec:syntax} and \ref{sec:metamodules}, but because it has been thoroughly studied in the literature, we will defer to prior work both here and in the dissertation for the formal details.% Instead, we will consider it abstractly where it interacts with the core language.%\todo{reminder that we are not going to re-specify the module system}
+
+The core language is where Verse diverges more radically from other languages, in that it is organized much like the first stage of a type-directed compiler (e.g.
+the TIL compiler for Standard ML \cite{tarditi+:til-OLD}), consisting of a user-facing \emph{typed external language} (EL) specified by type-directed translation to a minimal \emph{typed internal language} (IL).
+The main judgements in the specification of the EL take the following form  (omitting various contexts for now):\footnote{Note that we use the terms ``expression'' and ``type'' without qualification to refer to  external expressions and types, respectively.
+}
 \\[1ex]
@@
 For our purposes, we will keep things simple by using the strictly evaluated polymorphic lambda calculus with nullary and binary product and sum types and recursive types, which we assume the reader has familiarity with and for which all the necessary metatheorems are well-established (we follow \cite{pfpl}).
-The main judgements in the static semantics of the IL (again omitting contexts for now) take the following essentially standard form:
+The main judgements in the static semantics of the IL (again omitting contexts for now) take the following standard form:
 \\[1ex]
@@
 Verse, like most major contemporary programming languages, specifies a textual concrete syntax.
-We have chosen to specify a layout-sensitive textual concrete syntax (i.e.
+%We have chosen to specify a layout-sensitive textual concrete syntax (i.e.
 newlines and indentation are not ignored).
@@
 %For constructs that have an obvious analog in ML, we will omit a detailed explanation.
-
-A programming language's concrete syntax serves as its human-facing user interface, so it is common practice to build in  derived forms (colloquially, \emph{syntactic sugar}) that capture common idioms more concisely or naturally.
+A language's concrete syntax serves as its programmer-facing user interface, so it is common practice to build in  derived forms (colloquially, \emph{syntactic sugar}) that capture common idioms more concisely or naturally.
 % (i.e.
@@
 \begin{lstlisting}[numbers=none]
-casetype Pattern 
+casetype Pattern {
     Empty
-  Str of string
-  Seq of Pattern * Pattern
-  Or of Pattern * Pattern
-  Star of Pattern
-  Group of Pattern
+  | Str of string
+  | Seq of Pattern * Pattern
+  | Or of Pattern * Pattern
+  | Star of Pattern
+  | Group of Pattern
+}
 \end{lstlisting}
@@
 \begin{lstlisting}[deletekeywords={case},numbers=none]
-signature PATTERN
+signature PATTERN = sig 
   type t
@@
     'a)
+end
 \end{lstlisting}
@@
 \begin{lstlisting}[numbers=none]
-module PatternUtil(P : PATTERN)
-  fun parse(s : string) : P.t
-    (* ... pattern parser here ... *)
+module PatternUtil(P : PATTERN) = mod 
+  fun parse(s : string) : P.t = (* ... pattern parser here ... *)
+end 
 \end{lstlisting}
@@
 \begin{lstlisting}[numbers=none]
-case moon_phase
+case(moon_phase) {
     Full => pattern "SSTR(GCESTR" (* malformedness not statically detected *)
-  _ => (* ... *)
+  | _ => (* ... *)
+}
 \end{lstlisting}
@@
 Statically proving that this condition holds throughout a program is wildly difficult to automate in general, because it involves reasoning about arbitrary dynamic behavior.
-In the example above, we must know that the variable \lstinline{pattern} is equivalent to the function \lstinline{PU.pattern}.
-Moreover, if the argument were not written literally, e.g.
-we had written \lstinline{strconcat "SSTR(GESTR" "SSTRCESTR"}, we would need to be able to establish that this is equivalent to the string above.
-It is simple to come up with arbitrarily more complex examples that thwart any putative decision procedure, particularly for patterns that are dynamically constructed based on input to the program (further discussed below).
+In the example above, we must know that the variable \lstinline{pattern} binds the function \lstinline{PU.pattern}.
+If the argument had not been written literally but rather computed, e.g.
+if we had written \lstinline{"SSTR(GESTR" ^ "SSTRCESTR"} where \lstinline{^} is the string concatenation function applied in the infix style, we would need to be able to establish that this expression is equivalent to the string above.
+For patterns that are dynamically constructed based on input to the program (further discussed below), the problem is harder still.
 
@@
 \begin{lstlisting}[numbers=none,escapechar=|]
-fun example_bad(name : string)
+fun example_bad(name : string) => 
   pattern (name ^ "SSTR: \\d\\d\\d-\\d\\d-\\d\\d\\d\\dESTR")
@@
 \begin{lstlisting}[numbers=none]
-fun example_fixed(name : string)
+fun example_fixed(name : string) => 
   P.Seq(P.Str(name), P.Seq(pattern "SSTR: ESTR", ssn)) (* ssn as above *)
@@
 
-	\subsection{Existing Approaches}\label{sec:syntax-existing} The difficulties above are common whenever one uses  strings transiently to construct more richly structured data, merely for the sake of syntactic convenience.
+	\subsection{Existing Approaches}\label{sec:syntax-existing} The difficulties described above are common whenever strings are parsed dynamically to address the issue of syntactic cost (a static issue).
 %Strings are, simply put, not ideally suited for this task.
-	This key observation has motivated a substantial quantity of research on reducing the need for run-time string parsing \cite{Bravenboer:2007:PIA:1289971.1289975}.
-Existing approaches can be classified in  one of two ways: as being based on \emph{syntax extension} or \emph{static term rewriting}.
+	This has motivated a substantial quantity of research on reducing this use of run-time string parsing \cite{Bravenboer:2007:PIA:1289971.1289975}.
+The existing approaches can be classified in  one of two ways: as being based on \emph{syntax extension} or \emph{static term rewriting}.
 
 \subsubsection{Syntax Extension}\label{sec:syntax-extension}
-One approach to address the problems above would be to leverage a system that allows us to directly extend the syntax of our language with new derived syntactic forms.% for patterns.
-
-The simplest such systems are those where each new syntactic form is described by a single rewrite rule.
-For example, Gallina, the external language of the Coq proof assistant, provides such a system \cite{Coq:manual}.
-A formal account of systems like these has been developed by Griffin \cite{5134}.
+One approach we might consider is to directly extend the syntax of our language with new derived syntactic forms.% for patterns.
+
+The simplest such syntax extension systems are those where each new syntactic form is defined by a single rewrite rule.
+For example, Gallina, the ``external language'' of the Coq proof assistant, supports such extensions \cite{Coq:manual}.
+A formal account of such a system has been developed by Griffin \cite{5134}.
 Unfortunately, these systems are not flexible enough to allow us to express pattern syntax in the conventional manner.
 For example, sequences of characters  can only be parsed as identifiers using these systems, rather than as characters in a pattern.
-Syntax extension systems based on context-free grammars like  Sugar* \cite{erdweg2013framework}, Camlp4 \cite{ocaml-manual} and many others are more expressive, however, and thus would allow us to directly introduce pattern syntax into our core language's grammar, perhaps like this:
+Syntax extension systems based on context-free grammars like  Sugar* \cite{erdweg2013framework}, Camlp4 \cite{ocaml-manual} and many others are more expressive, and would allow us to directly introduce pattern syntax into our core language's grammar, perhaps like this:
 \begin{lstlisting}[numbers=none]
@@
 So if another library provider used similar syntax for a different implementation of regular expressions, or for some other unrelated abstraction, then a client could not simultaneously use both libraries in the same scope.
-Properly considered, every combination of extensions introduced by one of these mechanisms creates a \emph{de facto} derived syntactic dialect of our language.
+Properly considered, every combination of extensions introduced using one of these mechanisms creates a \emph{de facto} derived syntactic dialect of our language.
 % Resolving such parsing amibiguities is left to each client of the library.
@@
 
-Putting this issue aside, we must now decide how our newly introduced derived forms elaborate to forms in our base grammar.
-This is tricky because we have defined a modular encoding of patterns -- which particular module should the elaboration logic use?
+Putting this aside, we must now decide how our newly introduced forms elaborate to forms in our base grammar.
+This is tricky because we have defined a modular encoding of patterns -- which particular module should the elaboration use?
 Clearly, simply assuming that some module identified as \lstinline{P} satisying \lstinline{PATTERN} is in scope is a brittle solution.
-Indeed, we should expect that the extension mechanism actively prevents such capture of specific variable names to ensure that variables (including module variables) can be freely renamed.
-Such a \emph{hygiene discipline} is well-understood when performing term-to-term rewriting (discussed below) or in simple rewrite systems like those found in Coq.
+In fact, we should expect that the extension mechanism actively prevents such capture of specific variable names to ensure that variables (including module variables) can be freely renamed.
+Such a \emph{hygiene discipline} is well-understood only when performing term-to-term rewriting (discussed below) or in simple rewrite systems like those found in Coq.
 For more flexible mechanisms, the issue is a topic of ongoing research (none of the grammar-based mechanisms described above enforce hygiene).
@@
 \begin{lstlisting}[numbers=none,escapechar=|]
-fun example_syn(name : string)
+fun example_syn(name : string) => 
   edu_cmu_verse_rx_pattern P /@name: %ssn/
@@
 %The body of this function elaborates to the body of \lstinline{example_fixed} as shown above. 
-Had we mistakenly used the pattern splicing syntax, writing \lstinline{%name}, rather than the string splicing syntax, \lstinline{@name}, we would get a static type error, rather than the  silent injection  vulnerability that we discussed earlier.
+Had we mistakenly used the pattern splicing syntax, writing \lstinline{%name}, rather than the string splicing syntax, \lstinline{@name}, we would encounter only a static type error, rather than the  silent injection  vulnerability that we discussed earlier.
 
@@
 
-For patterns that are constructed compositionally, we need to get more creative.
+For patterns that are constructed compositionally, we would need to get more creative.
 For example, we might repurpose the infix operators that the language normally uses for other purposes to support string and pattern splicing as follows:
@@
 \begin{lstlisting}[numbers=none,escapechar=|]
-fun example_macro(name : string)
+fun example_macro(name : string) => 
   pattern P (name ^ "SSTR: ESTR" + ssn)
@@
 While this does not leave us with syntax that is quite as clean as would be possible with a  na\"ive syntax extension,  it does address many of the problems we discussed above: there cannot be syntactic conflicts (because the syntax is not extended), we can define macro abbreviations because macros are integrated into the language, there is a hygiene discipline that guarantees that the elaboration will not capture variables inadvertently, and by using a typed macro system, there is a typing discipline (i.e.
-creativewe need not examine the elaboration directly to know what type it must have).
+we need not examine the elaboration directly to know what type it must have).
 
@@
 \subsection{Contributions}\label{sec:syntax-contributions}
-We propose a new primitive, the \textbf{typed syntax macro} (TSM), that combines the flexibility of syntax extensions with the reasoning guarantees of statically-typed macros, and show how to integrate it into a language with an ML-style module system, addressing all of the problems described thusfar.
+We propose a new primitive, the \textbf{typed syntax macro} (TSM), that combines the flexibility of syntax extensions with the reasoning guarantees of statically-typed macros, and show how to integrate it into a language with an ML-style module system, addressing all of the brought up thusfar.
 We then introduce \textbf{type-specific languages} (TSLs), which leverage local type inference to control TSM dispatch, further decreasing syntactic cost (which is, of course, the entire purpose of this exercise).
-The result is that derived syntax defined by library providers using TSMs and TSLs  approaches the cost of derived syntax built in primitively by a language designer (using a na\"ive tool like Camlp4).
+The result is that derived syntax defined by library providers using TSMs and TSLs  approaches the cost of derived syntax built in primitively by a language designer (perhaps using a na\"ive tool like Camlp4).
 
@@
 %A typed syntax macro is invoked by applying it to a \emph{delimited form}, which can contain  arbitrary syntax in its \emph{body}.  
-Consider the following concrete external expression:
+Let us consider the following concrete external expression:
 \begin{lstlisting}[numbers=none]
@@
 \begin{lstlisting}[numbers=none]
-syntax pattern(P : PATTERN) at P.t
+syntax pattern(P : PATTERN) at P.t => 
   fn (ps : ParseStream) => (* pattern parser here *)
@@
 the TSM can be used with \emph{any} module matching signature \lstinline{PATTERN}.
-The module parameter is bound in the scope of the type clause \lstinline{at P.t}.
-This clause  specifies that all elaborations successfully generated by this TSM will be of type \lstinline{P.t}.
-Elaborations are generated by the static action of the parse function defined in the indented block on the next line.
-The parse function must be of type \lstinline{ParseStream -> Exp}, where the type \lstinline{ParseStream} gives the function access to the {body} of the delimited form (in blue above) and the type \lstinline{Exp} is a case type that encodes the abstract syntax of the base language (i.e.
-the Verse EL).
+The module parameter is used in the type annotation \lstinline{at P.t}, which specifies that all elaborations successfully generated by applying this TSM to a module parameter \lstinline{P} will be of type \lstinline{P.t}.
+Elaborations are generated by the static action of the parse function defined in the indented block on the second line.
+The parse function must be of type \lstinline{ParseStream -> Exp}.
 Both types are defined in the Verse \emph{prelude}, which is a set of definitions available ambiently.
+The type \lstinline{ParseStream} gives the function access to the {body} of the delimited form (in blue above) and the type \lstinline{Exp} is a case type that encodes the abstract syntax of external expressions.
 
@@
 \begin{lstlisting}[numbers=none]
-fun example_tsm(name: string)
+fun example_tsm(name: string) => 
   pat /SURL@EURLnameSURL: %EURLssn/
 \end{lstlisting}
-The hygiene mechanism ensures that only those portions of the elaboration derived from such spliced expressions can refer to variables at the site of use, preventing inadvertent variable capture by the elaboration.
-Put another way, the generated elaboration must be context-independent.
-
-\paragraph{Formalization} To give a flavor of the formal specification underlying TSMs, let us look at the rule for handling TSM invocation in a synthetic position: 
+A hygiene mechanism ensures that only portions of the generated elaboration derived from such spliced expressions can refer to the variables at the use site, preventing inadvertent variable capture by the elaboration.
+Put another way, the  elaboration logic must be valid in any context.
+
+\paragraph{Formalization} To give a flavor of the formal specification underlying TSMs, let us look at the rule for handling TSM application in a synthetic position: 
 \begin{mathpar}
@@
 }{
-    \Delta_\text{out}; \Gamma_\text{out}; \Delta; \Gamma \vdash \mathbf{invoketsm}(s; body) \Rightarrow \sigma \leadsto \iota
+    \Delta_\text{out}; \Gamma_\text{out}; \Delta; \Gamma \vdash \mathbf{aptsm}(s; body) \Rightarrow \sigma \leadsto \iota
 }
@@
 they omit various contexts relevant only to other Verse primitives.
-We include only \emph{typing contexts}, $\Gamma$, which map variables to types, and \emph{type abstraction contexts}, $\Delta$, which track universally quantified type variables (both in the standard way).
+We include only \emph{typing contexts}, $\Gamma$, which map variables to types, and \emph{type abstraction contexts}, $\Delta$, which track universally quantified type variables, both in the standard way.
 The premises have the following meanings:
 \begin{enumerate}
-\item The first premise can be pronounced ``Under $\Delta$, TSM expression $s$ is a TSM at type $\sigma$ with parser $\iota_\text{parser}$''.
-We elide the details of issues like module parameter application here for simplicity.
+\item The first premise can be pronounced ``TSM expression $s$ is a TSM at type $\sigma$ under $\Delta$ with parser $\iota_\text{parser}$''.
+This judgement would handle the mechanisms of module parameter application, but we must elide the details here (a complete specification of this judgement represents the main avenue for remaining work).
 % Application of a parameterized TSM to its parameters is handled by this judgement (not shown).
@@
 \begin{lstlisting}[numbers=none]
-module P : PATTERN with syntax pattern at t
+module P : PATTERN with syntax pattern(P) at t = mod 
   type t = (* ... *)
   (* ... *)
+end 
 \end{lstlisting}
@@
 We have described and given a more detailed formal specification of TSMs in a recently published paper \cite{sac15}, and TSLs in a paper last year \cite{TSLs}, both in the context of the Wyvern language.
-At the time of publication, Wyvern did not specify a module system, so in the context of Verse, there are some changes that need to be made in the formalization to handle module parameters.
-We plan to complete this in the course of writing the dissertation.
+The mechanics of elaboration generation, encoding and decoding and hygienic elaboration validation have all been satisfyingly specified in this context.
+
+Wyvern does not specify an ML-style module system (type declarations provide a form of type generativity), so in the context of Verse, there are some changes that need to be made in the formalization to handle module parameters, as mentioned earlier.
+We plan to complete this in the course of writing the dissertation, and do not anticipate it requiring more than 1-2 weeks of effort over the summer.
 
@@
 In contrast to other languages, which build in constructs like case types (shown briefly at the beginning of Sec.
-\ref{sec:examples}), tuples, records, objects and so on primitively, in Verse these can all be expressed as modes of use of \emph{metamodules} (which are distinct from, though conceptually related to, \emph{modules}).
-In fact, only polymorphic function types are built primitively into the Verse EL.
+\ref{sec:examples}), tuples, records, objects and so on primitively, in Verse these can all be expressed as modes of use of \emph{metamodules} (which are distinct from, though conceptually related to, {modules}, as we will see).
+Only polymorphic function types are built primitively into the Verse EL.
 
@@
 \begin{lstlisting}[numbers=none]
-let type Paper = ltuple {SPCT
+let type Paper = ltuplex {SPCT
    title : EPCTrstring /SURL.+EURL/
@@
 \end{lstlisting}
-The \lstinline{let type} construction defines a synonym, \lstinline{Paper}, for a type constructed by applying the \emph{type constructor} (or more concisely, \emph{tycon}) \lstinline{ltuple} to an ordered finite mapping from labels to types, delimited here by curly braces.
-Its first component is labeled \lstinline{title} and specifies the regular string type \lstinline{rstring /SURL.*EURL/}.
-Regular string types are constructed by applying the tycon \lstinline{rstring} to a regular expression pattern,  written literally here.
-They classify values that behave like strings in the corresponding regular language.
-So in this case, we have specified that paper titles must be non-empty strings.
-The second component of our labeled product type is labeled \lstinline{conf} and specifies a regular string type with two parenthesized groups, corresponding to the conference abbreviation and year.
+The \lstinline{let type} construction defines a synonym, \lstinline{Paper}, for a type constructed by applying the \emph{type constructor} (or \emph{tycon}) \lstinline{ltuplex} to a statically valued ordered finite mapping from labels to types, here written using a conventional concrete syntax within curly braces (this is a mode of use of TSLs, as we will see).
+Its first component is labeled \lstinline{title} and specifies the regular string type \lstinline{rstring /SURL.+EURL/}.
+Regular string types are constructed by applying the tycon \lstinline{rstring} to a regular expression pattern, again written using a conventional concrete syntax.
+Regular string types classify values that behave like strings in the corresponding regular language, here the language of non-empty strings.
+The second component of type \lstinline{Paper} is labeled \lstinline{conf} and specifies a regular string type with two parenthesized groups, corresponding to a conference abbreviation and year.
 
@@
 \end{lstlisting}
-Here, \lstinline{#} is an \emph{operator constructor} (or \emph{opcon}) parameterized by a static label, written literally here as \lstinline{<SNATconfENAT>}.
+Here, \lstinline{#} identifies an \emph{operator constructor} (or \emph{opcon}) parameterized by a static label, written literally here as \lstinline{<SNATconfENAT>}.
 The resulting \emph{operator}, \lstinline{# <SNATconfENAT>}, takes one \emph{argument}, \lstinline{exmpl}.
 
-We can also project captured groups out of a regular string using the \lstinline{#group} operator constructor, which is parameterized by a natural number referring to the group index:
+Given a value of regular string type like this, we can project out captured groups using the \lstinline{#group} operator constructor, which is parameterized by a natural number referring to the group index:
 \begin{lstlisting}[numbers=none]
@@
 \end{lstlisting}
-Note that the variable \lstinline{exmpl_conf_name} has type \lstinline{rstring /SURL[A-Z]+EURL/}.
-
-Both labeled tuples and regular strings support a number of other basic operations.
+The variable \lstinline{exmpl_conf_name} has type \lstinline{rstring /SURL[A-Z]+EURL/}.
+
+Our take on labeled tuples supports a number of other basic operations.
 For example, labeled tuples can be concatenated (with any common components updated with the value on the right) using the \lstinline{ltuple+} operator:
 \begin{lstlisting}[numbers=none]
-let a : lprodtype {SPCTauthors : EPCTlist(rstring /SURL.+EURL/)} = {["SUSHarry Q. BovikEUS"]}
+let a : ltuple {SPCTauthors : EPCTlist(rstring /SURL.+EURL/)} = {["SUSHarry Q. BovikEUS"]}
 let exmpl_paper_final = ltuple+ exmpl_paper a
@@
 \subsection{Existing Approaches}\label{sec:metamodules-related}
-\subsubsection{Labeled Product Types}
+\subsubsection{Labeled Tuple Types}
 Recall that Verse builds in only nullary and binary products into its internal language, so the type structure described above for labeled tuples  is not directly expressible using IL primitives.
@@
 For our example above:
-For example, we might define the following signature for our example above:
+For example, we might define the following signature for our example type \lstinline{Paper} from earlier:
 \begin{lstlisting}
@@
 We've simply taken all possible valid introductory and projection operators and expressed them as functions, moving label parameters into identifiers.
-Even with the minimal EL we are assuming, we could  implement this signature using a Church-style encoding.
-The details are straightforward and omitted.
-Alternatively, if we only slightly relax slightly our insistence on minimalism and add binary products to the EL, we could use them as well:
+Even with our minimal EL, we could  implement this signature using a Church-style encoding (the details are standard and omitted).
+Alternatively, if we only slightly relax  our insistence on minimalism and add binary products to the EL, we could use these as well:
 \begin{lstlisting}
@@
 
-One problem is that this requires a volume of boilerplate code quadratic in the number of components of our labeled tuple type.
-It is also not entirely satisfying syntactically from the client's perspective.
+There are several basic problems with this approach.
+First,  each labeled tuple type that a library provider wishes to encode in this way requires a volume of boilerplate code factorial in the number of components it has (because we must consider all permutations of label orderings as a distinct introductory function).
+It is also not satisfying syntactically from the client's perspective.
 To avoid these problems and recover the syntax used previously, we could define a TSM for every such signature, though this too involves much boilerplate.
 
-More fundamentally, this encoding is limited to expressing the basic introductory and projection operators.
-We did not express all possible valid operators that arise from the opcon \lstinline{ltuple-} as functions, because to do so, we would need to construct an encoding of every labeled tuple type that might arise from dropping one or more components.
+More fundamentally, we could only express the introductory and projection operators.
+We were not able to enumerate all possible operators that arise from the opcon \lstinline{ltuple-} as functions, because to do so, we would need to construct an encoding of every labeled tuple type that might arise from dropping one or more components.
 This would require an exponential volume of boilerplate code.
-Finally, there is no way to define \lstinline{ltuple+} as a function or a finite collection of functions, because there are infinitely many extensions of any particular labeled product type.% of this operator would need to be implemented by clients manually, defeating the purpose of the operator.
+Finally, there is simply no way to define \lstinline{ltuple+} as a finite collection of functions -- there are infinitely many extensions of any particular labeled product type.
+% of this operator would need to be implemented by clients manually, defeating the purpose of the operator.
 %For example, there are an arbitrary
@@
 In other words, this mapping does not preserve certain type disequalities between labeled tuple types.
-Lacking positional information, it is not possible to omit component labels and rely on positional information when introducing a value of this type.%, because this information is not maintained by the type system.
+Lacking positional information, it is not possible to omit component labels and rely on positional information when introducing a value of this type directly.%, because this information is not maintained by the type system.
 
@@
 
-If on the other hand we do care about preserving the type disequality mentioned, the workaround is even less elegant: we first need to define a record type with component labels tagged in some sufficiently unambiguous way by position, e.g.:
+If on the other hand we do care about preserving the type disequalities in question, the workaround is even less elegant: we first need to define a record type with component labels tagged in some sufficiently unambiguous way by position, e.g.:
 \begin{lstlisting}[numbers=none]
@@
 %This treats membership of a string in a specified regular language exclusively as a static verification condition.
-These systems typically examine refinement specifications supplied in comments:
+Such  systems typically rely on refinement specifications supplied in comments:
 \begin{lstlisting}[numbers=none]
@@
 Expressing concatenation uniformly using a single function or a finite set of functions is impossible, for the same reason that applied to \lstinline{ltuple+} above, i.e.
-we would need to define one such function for every pair of possible regular string types (of which there are infinitely many).
+there are infinitely many regular string types.
 
@@
 
-Metamodules introduce type and operator constructors parameterized by arbitrary \emph{static values} and metaprogrammatically express the logic governing their typechecking and  translation to the IL.
-Metamodules are classified by \emph{metasignatures}.
-For example, let us consider the following metasignature:
+\emph{Metamodules} introduce type and operator constructors parameterized by arbitrary \emph{static values} and metaprogrammatically express the logic governing their typechecking and  translation to the IL.
+\emph{Metasignatures} classify metamodules.
+Let us begin by considering the following metasignature, which classifies metamodules that express the semantics of labeled tuples as described above:
+\begin{figure}
 \begin{lstlisting}
-metasignature LTUPLE
-  tycon c of LblTyMap
-  ana opcon intro_unlabeled
-  ana opcon intro_labeled of LblList
+metasignature LTUPLEX
+  tycon c of LblTyOrdMap
+  ana opcon intro_unlabeled (*of unit*)
+  ana opcon intro_labeled of list(Lbl)
   syn opcon # of Lbl
-syntax ltpl(L ~ LTUPLE, idx : LblTyList) at L.c(idx)
-  ...
-metamodule LTuple ~ LTUPLE with syntax ltpl for c
-  tycon c(idx : LblTyList) ~> ...
-  ana opcon intro_unlabeled(ty : Ty, args : List(Arg)) ~> ...
-  ana opcon intro_labeled(lbls : List(Lbl), ty : Ty, args : List(Arg)) ~> ...
-  syn opcon #(lbl, args) ~> ...
-let tycon ltuple = LTuple.c
-let opcon # = LTuple.#
-\end{lstlisting}
+  syn opcon + (*of unit*)
+  syn opcon - of Lbl
+
+syntax ltpl(L ~ LTUPLEX, param : LblTyOrdMap) at L.c(param)
+  fn ps => (* ... intro syntax ... *)
+
+metamodule Ltuplex ~ LTUPLEX with syntax ltpl(Ltuplex) at c
+  (* translate labeled tuple types to nested products *)
+  tycon c(param : LblTyOrdMap) ~> (fold (list_of_ltom param)
+    `SCSSunitECSS`
+    (fn ((lbl, ty), cur_trans) => `SCSStrans(ECSStySCSS) * %ECSScur_trans`)
+  )
+
+  (* translate intro operator to nested pairs *)
+  ana opcon intro_unlabeled(ty : Ty, param : unit, args : list(Arg)) ~> (
+    tycase(ty) {
+      c(param) => (fold (zipExact (list_of_ltom param) args)
+      	`()`
+      	(fn (((lbl, ty), arg), cur_trans) => `(%{ana(arg)}, %cur_trans)`)
+      )
+      _ => raise TyErr("...")
+    }
+  )
+
+  (* same as intro_unlabeled but reorder first *)
+  ana opcon intro_labeled(ty : Ty, param : list(Lbl), args : list(Arg)) ~> 
+    (* ... *)
+
+  (* generate the appropriate n-fold projection out of nested pairs *)
+  syn opcon #(param : Lbl, args : list(Arg)) ~> 
+    (* ... *)
+
+  (* generate the new type ..... *)
+  syn opcon +(param : unit, args : list(Arg)) ~> 
+    (* ... *)
+  syn opcon -(param : Lbl, args : list(Arg)) ~> 
+    (* ... *)
+ (* synonyms used in Sec. 5.1 *)
+ let tycon ltuplex = LTupleX.c
+ let opcon # = LTupleX.#
+ let opcon ltuple+ = LTupleX.+ 
+ let opcon ltuple- = LTupleX.-
+
+\end{lstlisting}
+\caption{Using metamodules to define the type structure of labeled tuple types.}
+\label{fig:ltuplex}
+\end{figure}
+
+On line 2, we declare a tycon, identifying it as \lstinline{c}, and specify that it is parameterized by static values of type \lstinline{LblTyOrdMap}, which we assume has been defined as the type of ordered finite mappings from labels, which are values of type \lstinline{Lbl}, to types, which can be manipulated as static values of type \lstinline{Ty}.
+
+Let us assume a metamodule \lstinline{LTupleX} that implements this metasignature,
 

### 2015-04-19 thesis-proposal/problem.tex
updates

@@
 Instead, its users are able to modularly  express a wide variety of useful constructs in libraries, using the comparatively small set of primitives that the language builds in.
-Functional programming languages like Standard ML (SML) and Haskell descend directly from  elegant typed lambda calculi, diverging mainly in that they consider  various human factors in addition to fundamental metatheoretic issues in their design.
-For example, they both build in record types, generalizing the nullary and binary product types that are more common in simpler calculi, because labeled components are cognitively useful.
-Similarly, they both build in derived  syntax for the standard list datatype.% concision and readability.
+Functional programming languages like Standard ML (SML) and Haskell descend directly from typed lambda calculi, diverging mainly in that they consider  various usability criteria in addition to fundamental metatheoretic issues in their design.
+For example, they build in record types, generalizing the nullary and binary product types more common in simpler calculi, because labeled components are cognitively useful to programmers.
+Similarly, they both build in derived syntax for the list datatype.% concision and readability.
 %Minimal typed lambda calculi play a central role in the study of programming languages, and serve as direct inspiration for  functional programming languages like ML.
@@
 
-Ideally, the transition from  a minimal calculus to a pragmatic ``full-scale'' programming language would require only a limited number of primitive ``embellishments'' like these.
-The community around the language could then direct its efforts almost exclusively toward the development of useful libraries.
-Unfortunately, a single language design that achieves this ideal has proven elusive, as evidenced by the diverse array of {dialects} that continue to proliferate around all major contemporary languages, functional languages included.
+Ideally, this transition from a minimal calculus to a pragmatic ``full-scale'' programming language would require only a limited number of primitive ``embellishments'' like these.
+The community around the language would then be able to direct its efforts exclusively toward the development of useful libraries.
+Unfortunately, a stable language design that achieves this ideal has yet to emerge, as evidenced by the diverse array of {dialects} that continue to proliferate around all major contemporary languages, functional languages like ML included.
 % For example, let us consider the many dialects of ML.
@@
 %Tools for  constructing new ``domain-specific'' dialects also continue to proliferate. % The situation is similar across all major language lineages. 
-In fact, tools that make constructing ``domain-specific'' language dialects easier appear to be increasingly popular. 
-{This calls for an investigation}: why are well-informed programmers and researchers so often unable to satisfyingly express the constructs that they need in libraries, as modes of use of the ``general-purpose'' functional primitives already available in major contemporary languages today?
+Indeed, tools that assist in the construction of ``domain-specific'' language dialects  seem only to be increasing in popularity. 
+{This calls for an investigation}: why are well-informed programmers and researchers so often unable to satisfyingly express the constructs that they need in libraries, as modes of use of the ``general-purpose'' functional primitives already available in major languages today?
 %Let us investigate why this is.
@@
 
-Perhaps the most common  reason may simply be that the \emph{syntactic cost} of expressing a construct of interest using contemporary general-purpose primitives is not always ideal.
+Perhaps the most common  reason may be that the \emph{syntactic cost} of expressing a construct of interest in terms of contemporary general-purpose primitives is not always ideal.
 %Syntactic cost is often assessed qualitatively \cite{green1996usability}, though quantitative metrics can be defined.
-For example,  we consider regular expression patterns expressed using abstract data types (in ML, a mode of use of the module system) in Sec.
-\ref{sec:syntax}.
-This  gives  us precisely the semantics that we seek to express, but syntactically, we argue that the usual approach of using strings to introduce patterns leaves much to be desired.
-To avoid this, we are tempted to introduce derived syntax specific to our library by constructing a \emph{derived syntactic dialect}, i.e.
+For example,  we will consider regular expression patterns expressed using abstract data types (in ML, a mode of use of the module system) in Section \ref{sec:syntax}.
+This  will allow  us to express  precisely the semantics that we seek, but syntactically,  the usual approach of using run-time string parsing whenever we introduce a pattern will leave something to be desired.
+To avoid run-time string parsing, we will be tempted to introduce derived syntactic forms specific to our library, thereby constructing a \emph{derived syntactic dialect}, i.e.
 a dialect specified by context-free elaboration to the existing language.
-This is far from an isolated example.
-For example, Ur/Web is a  derived syntactic dialect of Ur (itself descended from ML) that builds in derived syntax for SQL queries, HTML elements and other datatypes common in web programming \cite{conf/popl/Chlipala15}.
+Such derived syntactic dialects are not uncommon.
+For example Ur/Web is a  derived syntactic dialect of Ur (itself descended from ML) that builds in derived syntax for SQL queries, HTML elements and other datatypes common in web programming \cite{conf/popl/Chlipala15}.
 %In fact, nearly all major programming languages primitively build in derived syntax for constructs that can otherwise be expressed in the standard library, e.g.
 derived list syntax in SML \cite{harper1997programming,mthm97-for-dart}.
-Tools like Camlp4 \cite{ocaml-manual} and Sugar* \cite{erdweg2011sugarj,erdweg2013framework} (which we will return to later) are designed to lower the engineering costs of constructing derived syntactic dialects, contributing to their proliferation.% contributing to their proliferation.
+Tools like Camlp4 \cite{ocaml-manual} and Sugar* \cite{erdweg2011sugarj,erdweg2013framework} are designed to lower the engineering costs of constructing derived syntactic dialects, contributing to their continued proliferation (we will consider them further in Sec.
+\ref{sec:syntax}).
+% contributing to their proliferation.
 
@@
 
-More advanced dialects introduce  new type structure, going beyond what is possible by context-free elaboration to an existing language.
-For example, as mentioned above, SML goes beyond nullary and binary product types to expose record types.
-%Note however that the dynamic semantics of record types can, of course, be expressed by translation to a minimal calculus.
-Other dialects have explored  ``record-like'' primitives that support functional update and extension operators, width and depth coercions (sometimes implicit)%\cite{Cardelli:1984:SMI:1096.1098}
-, mutable components, methods, prototypic dispatch and various other embellishments.
-Ocaml builds in an object system, polymorphic variants and  conveniences like logic for typechecking operations that use format strings, e.g.
-$\mathtt{sprintf}$ \cite{ocaml-manual}.
+More advanced dialects define new static type structure, going beyond what is possible by context-free elaboration to the existing language.
+Record types themselves are perhaps  the simplest example -- they cannot be expressed by context-free elaboration to a language with only nullary and binary products.
+Various dialects of ML and Haskell have explored  ``record-like'' primitives that go further, supporting functional update and extension operators, width and depth coercions (sometimes implicit)%\cite{Cardelli:1984:SMI:1096.1098}
+, methods, prototypic dispatch and various other semantic ``embellishments'' that cannot be expressed by context-free elaboration to a language with standard record types (we will detail an example in Sec.
+\ref{sec:metamodules}).
+The Ocaml dialect of ML builds in ``polymorphic variants'', ``open datatypes'' and  logic for typechecking operations that use format strings like $\mathtt{sprintf}$ \cite{ocaml-manual}.
 ReactiveML builds in primitives for functional reactive programming \cite{mandel2005reactiveml}.
@@
 Alice ML builds in futures and promises \cite{AliceLookingGlass}.
-MLj builds in essentially the entire Java programming language, to support typesafe interoperation with existing Java programs \cite{Benton:1999:IWW:317636.317791}.
-Indeed, perusal of the proceedings of any major programming languages conference will typically yield several more examples.
-Tools like proof assistants and logical frameworks support specifying and reasoning metatheoretically about dialects like these, and tools like compiler generators and language workbenches lower the cost of implementing them, again contributing to their continued proliferation.
+MLj builds in the entire Java programming language, to support typesafe interoperation with Java libraries \cite{Benton:1999:IWW:317636.317791}.
+Tools like proof assistants and logical frameworks support specifying and reasoning metatheoretically about dialects like these, and tools like compiler generators and language workbenches lower the cost of implementing them, contributing to their  proliferation.
 
@@
 given two dialects that are specified by judgements of arbitrary form, it is not clear what it would even mean to combine them.
-Even when the dialects are specified using some formalism that does provide a means to na\"ively combine two dialects, there is generally no guarantee that the composition will conserve important syntactic and semantic properties.
+Even when the dialects are specified using some formalism that does define what it means to combine two dialects, there is generally no guarantee that the composition will conserve important syntactic and semantic properties.
 %There is no well-defined mechanism for constructing such a ``combined language'' in general.
 For example, consider two derived syntactic dialects, one building in derived syntax for JSON (a popular data interchange format), the other using a similar syntax for finite mappings of some other type.
-Though each is known to have an unambiguous concrete syntax in isolation, when their syntax is na\"ively  combined (e.g.
+Though each is known to have an unambiguous concrete syntax in isolation, when their grammars are na\"ively  combined (e.g.
 by Camlp4), ambiguities arise.
@@
 
-Dialects do have an indirect impact on large-scale software development because the language designers that control comparatively popular languages, like Ocaml and Scala, can be convinced to incorporate  ideas from dialects  into backwards compatible language revisions.
+Dialects do, however, have a more indirect impact on large-scale software development in that the language designers that control comparatively popular languages, like Ocaml and Scala, can be convinced to incorporate  ideas from dialects  into backwards compatible language revisions.
 %These decisions are increasingly influenced by community processes, e.g.
@@
 However, this has come at a cost: these languages have ballooned in size.
-And yet, because there appears to be a ``long tail'' of potentially useful primitives (for derived syntax, our group has  gathered initial data speaking to this \cite{TSLs}),  primitives that are only situationally useful, or that trigger aesthetic disagreements between different factions of the community, are still quite often left languishing in ``toy'' dialects.
+This is not a scalable approach -- as suggested by the diversity of examples above, there is a ``long tail'' of potentially useful such language extensions (for derived syntax, our group has  gathered initial data speaking to this \cite{TSLs}).
+%rimitives that are only situationally useful, or that trigger aesthetic disagreements between different factions of the community, are still quite often left languishing in impractical ``toy'' dialects.
 Recalling the words of  Reynolds, which are as relevant today as they were nearly half a century ago \cite{Reynolds70}:%Because there is no data about how useful a construct is in practice until it is included in a language like this, decisions about which constructs to include are often informed by little more than intuition.
@@
 Encouragingly, primitives that have such a ``slimming'' effect on a language do occasionally arise.
-For example, Ocaml in a recent revision added support for  ``generalized algebraic data types'' (GADTs), based on research on guarded recursive datatype constructors \cite{XiCheChe03}.
-Using GADTs, Ocaml was able to move some of the machinery for typechecking operations that use format strings, like \texttt{sprintf}, out of the language and into the standard library (unfortunately, some machinery remains built in).
-Our broad aim is to introduce a new functional programming language called Verse\footnote{We distinguish Verse from Wyvern, which is the language referred to in prior publications about some of the work being proposed here, because Wyvern is a group effort evolving independently in some important ways.} that takes even more dramatic steps down this second path.
+For example, a recent revision of Ocaml added support for  ``generalized algebraic data types'' (GADTs), based on research on guarded recursive datatype constructors \cite{XiCheChe03}.
+Using GADTs, Ocaml was able to move some of the machinery for typechecking operations that use format strings, like \texttt{sprintf}, out of the language and into the standard library (but some machinery remains built in).
+Our broad aim in the work being proposed is to introduce a new functional programming language called Verse\footnote{We distinguish Verse from Wyvern, which is the language referred to in prior publications about some of the work being proposed here, because Wyvern is a group effort evolving independently in some important ways.} that takes  more dramatic steps down this second path by allowing library providers to express new derived syntax and type structure of their own design, in a safe and modularly composable manner.
+
 %Similarly, it recently introduced ``open datatypes'', which subsume its previous more specialized exception type, and captures many use cases for .
@@
 
-Verse has a module system taken directly from SML, but its core language is specified by a \emph{type-directed translation semantics} targeting a minimal typed lambda calculus, called the Verse \emph{internal language}, introduced in Sec.
+Verse features a module system taken directly from SML, but its core language is organized in a novel manner as a \emph{type-directed translation semantics} targeting a minimal typed lambda calculus called the Verse \emph{internal language} (IL).
+We formally introduce Verse's core semantics in Sec.
 \ref{sec:verse}.
-In lieu of typical \emph{external} (i.e.
+In lieu of the typical \emph{external} (i.e.
 programmer-facing) derived syntax and  type structure, Verse introduces two novel primitive constructs:
@@
 list syntax as in SML or XML syntax as in Scala and Ur/Web, by giving library providers static control over the parsing and elaboration of delimited segments of concrete syntax.
+%We show how TSMs can be invoked explicitly by clients, or, implicitly based on local type inference.
 \item \textbf{Metamodules}, introduced in Sec.
-\ref{sec:metamodules}, subsume the need to primitively build in the type structure of constructs like records (and variants thereof) by giving library providers static hooks directly into the type-directed translation semantics governing the Verse external language.
-For example, we will discuss how a library provider can define a single metamodule that defines \emph{labeled tuple types} together with the operators that allow clients to introduce, project out of and concatenate values of these types.
-The internal language builds in only nullary and binary product types primitively.
+\ref{sec:metamodules}, subsume the need to primitively build in the type structure of constructs like records (and variants thereof),  labeled sums and other interesting constructs by giving library providers static hooks directly into the type-directed translation semantics.
+For example, a metamodule that implements the type structure of records would:
+\begin{enumerate}
+\item Introduce a type constructor \lstinline{record} parameterized by finite mappings from labels to types, defining its semantics by translation to (for example) unary and binary products (which are built in to the internal language); and
+\item Introduce operators used to work with records, minimally record introduction and elimination, but perhaps also various functional update operators, defining the logic governing their typechecking and translation to the IL directly.
+\end{enumerate}
+We will see direct analogies between modules and metamodules as we continue.
 \end{itemize} 
-Both TSMs and metamodules can be understood as \emph{metaprogramming} primitives, because they involve programmatic static code generation.
+Both TSMs and metamodules can be understood as \emph{metaprogramming} primitives, because their static semantics involves the invocation of user-defined functions.% static code generation.
 %We will  also introduce a simple variant of each of these primitives that leverages Verse's support for local type inference to further reduce syntactic cost in certain common situations.
 
-The key challenge in the design of these primitives will come in ensuring that they are metatheoretically well-behaved, given that they aim to give library providers decentralized control over aspects of Verse's concrete syntax and type structure that, in other languages, are under the monolithic control of the language designer.
+The key challenge in the design of these primitives will come in ensuring that they are metatheoretically well-behaved, given that they aim to decentralize control over aspects of the language's concrete syntax and type structure that, in other contemporary languages, are under the exclusive control of the language designer.
 If we are not careful, many of the problems discussed earlier as inherent to combining distinct language dialects could simply shift into the semantics of these primitives.\footnote{This is why languages  like Verse are often called \emph{extensible languages}, though this is somewhat of a misnomer.
@@
 \section{Summary of Judgements}\label{sec:verse}
-To situate ourselves within a formal framework, let us begin with a brief summary of how Verse is organized and specified. Verse consists of a \emph{module language} atop a \emph{core language}.
+To situate ourselves within a formal framework, let us begin with a brief summary of how Verse is organized and specified. %Verse consists of a \emph{module language} atop a \emph{core language}.
 
@@
 
-The core language is where Verse diverges more radically from other languages, in that it is organized much like the first stage of a type-directed compiler (e.g.
+The core language is organized much like the first stage of a type-directed compiler (e.g.
 the TIL compiler for Standard ML \cite{tarditi+:til-OLD}), consisting of a user-facing \emph{typed external language} (EL) specified by type-directed translation to a minimal \emph{typed internal language} (IL).
@@
 This is to  allow us to explicitly specify how Verse's \emph{local type inference} works, and in particular, how it interacts with the primitives that we aim to introduce.
-Like Scala, we intentionally do not attempt to support global type inference, primarily because it is not technically possible to do so (but also because we believe that compilers for languages that support only local type inference generate clearer error messages).
+Like Scala, we intentionally do not attempt to support global type inference.%, primarily because it is not technically compatible with the primitives we will introduce (but also because we believe that compilers for languages that support only local type inference generate clearer error messages).
 

### 2015-04-28 thesis-proposal/parsing.tex
Updated proposal
Updated poropsal

@@
 Objects can thus be thought of as recursive labeled products with support for simple methods (functions that are automatically given a self-reference) for convenience.
-We also add recursive labeled sum types, which we call \emph{case types}, that are quite similar to datatypes in ML.
+We also add recursive labeled sum types, which we call \emph{variant types}, that are quite similar to datatypes in ML.
 One can refer to the version of the language described in this thesis as \emph{TSL Wyvern}.
@@
 Note that the delimiters used to go from Wyvern to a TSL are controlled by Wyvern  while the TSL controls how to return to Wyvern.
-%\subsection{General-Purpose Notation for Object and Case Types}
+%\subsection{General-Purpose Notation for Object and variant types}
 %The general-purpose introductory form for object types is \li{new}.
@@
 Object types in TSL Wyvern are simple structural interfaces that constrain the signatures of fields and methods.
-The \emph{type ascription} around \li{new} checks that the object being introduced satisfies the signature of \li{SearchServer} (not shown).\todo{discuss case types}
+The \emph{type ascription} around \li{new} checks that the object being introduced satisfies the signature of \li{SearchServer} (not shown).\todo{discuss variant types}
 \subsection{Layout-Delimited Literals}
@@
 \ref{f-example}, we see a call to a function \li{serve} (not shown) which has type \li{HTML -> Unit}.
-Here, \li{HTML} is a user-defined \emph{case type}, having cases for each HTML tag as well as some other structures, like text nodes and sequencing.
+Here, \li{HTML} is a user-defined \emph{variant type}, having cases for each HTML tag as well as some other structures, like text nodes and sequencing.
 Declarations  of some of these cases can be seen on lines 2-6 of Fig.
 \ref{f-htmltype} (note that TSL Wyvern also includes simple product types for convenience, written \li{T1 * T2}).
-We could again use Wyvern's general-purpose introductory form for case types, e.g.
+We could again use Wyvern's general-purpose introductory form for variant types, e.g.
 \li{BodyElement((attrs, child))}.
@@
 \begin{lstlisting}[escapechar=$]
-casetype HTML 
+variant HTML 
   Empty
@@
 \vspace{-8px}
-\caption{A Wyvern case type with an associated TSL.}
+\caption{A Wyvern variant type with an associated TSL.}
 \label{f-htmltype}
@@
     val parser = (* parser generator *)
-casetype ParseResult
+variant ParseResult
   OK of Exp * ParseStream
@@
 \begin{lstlisting}[linewidth=.42\textwidth,firstnumber=10]
-casetype Exp 
+variant Exp 
   Var of ID
@@
 The \li{Parser} type is an object type declaring a single method, \li{parse}, that transforms a \li{ParseStream} extracted from a literal body to a Wyvern AST.
-An AST is a value of type \li{Exp}, a case type that encodes the abstract syntax of Wyvern expressions.
+An AST is a value of type \li{Exp}, a variant type that encodes the abstract syntax of Wyvern expressions.
 Fig.
@@
 This work was published at ECOOP 2014 (and received a Distinguished Paper Award) \cite{TSLs}.
-We have not formally shown how parameterized types work (e.g.
+We have not formally shown how parameterized and abstract types work (e.g.
 we can't yet implement list literal syntax uniformly as a library).

### 2015-04-28 thesis-proposal/problem.tex
Updated poropsal

@@
 Instead, its users are able to modularly  express a wide variety of useful constructs in libraries, using the comparatively small set of primitives that the language builds in.
-Functional programming languages like Standard ML (SML) and Haskell descend directly from typed lambda calculi, diverging mainly in that they consider  various usability criteria in addition to fundamental metatheoretic issues in their design.
-For example, they build in record types, generalizing the nullary and binary product types more common in simpler calculi, because labeled components are cognitively useful to programmers.
-Similarly, they both build in derived syntax for the list datatype.% concision and readability.
+Functional programming languages like Standard ML (SML), OCaml and Haskell descend from mathematically elegant typed lambda calculi, diverging mainly in that they consider various pragmatic issues alongside  fundamental metatheoretic issues in their design.
+For example, they build in record types, which generalize the nullary and binary product types more common in minimal calculi, because labeled components can be cognitively useful to humans.
+Similarly, they build in derived syntax for commonly used constructs like lists.% concision and readability.
 %Minimal typed lambda calculi play a central role in the study of programming languages, and serve as direct inspiration for  functional programming languages like ML.
@@
 
-Ideally, this transition from a minimal calculus to a pragmatic ``full-scale'' programming language would require only a limited number of primitive ``embellishments'' like these.
-The community around the language would then be able to direct its efforts exclusively toward the development of useful libraries.
-Unfortunately, a stable language design that achieves this ideal has yet to emerge, as evidenced by the diverse array of {dialects} that continue to proliferate around all major contemporary languages, functional languages like ML included.
+Ideally, a limited number of ``embellishments'' like these would suffice to transition from a minimal calculus to a practical ``full-scale'' programming language.
+The community around the resulting language would then be able to direct its efforts exclusively toward the development of useful libraries.
+Unfortunately, a stable language design that achieves this ideal has yet to emerge, as evidenced by the diverse array of {dialects} that continue to proliferate around all major contemporary languages.
 % For example, let us consider the many dialects of ML.
@@
 %Tools for  constructing new ``domain-specific'' dialects also continue to proliferate. % The situation is similar across all major language lineages. 
-Indeed, tools that assist in the construction of ``domain-specific'' language dialects  seem only to be increasing in popularity. 
-{This calls for an investigation}: why are well-informed programmers and researchers so often unable to satisfyingly express the constructs that they need in libraries, as modes of use of the ``general-purpose'' functional primitives already available in major languages today?
+Indeed, tools that assist with the construction of  ``domain-specific'' language dialects seem only to be increasingly popular. 
+{This calls for an investigation}: why are programmers and researchers still so often unable to satisfyingly express the constructs that they need in libraries, as modes of use of the ``general-purpose'' primitives already available in major functional languages?
 %Let us investigate why this is.
@@
 
-Perhaps the most common  reason may be that the \emph{syntactic cost} of expressing a construct of interest in terms of contemporary general-purpose primitives is not always ideal.
+Perhaps the most common reason may simply be that the \emph{syntactic cost} of expressing a construct of interest using contemporary general-purpose primitives is not always ideal.
+To decrease syntactic cost, library providers are often tempted to construct \emph{derived syntactic dialects}, i.e.
+dialects that are specified by context-free elaboration to the existing language.
+For example, Ur/Web is a  derived syntactic dialect of Ur (a language that descends from ML) that builds in derived syntax (colloquially, ``syntactic sugar'') for SQL queries, HTML elements and other datatypes used for web programming \cite{conf/popl/Chlipala15}.
 %Syntactic cost is often assessed qualitatively \cite{green1996usability}, though quantitative metrics can be defined.
-For example,  we will consider regular expression patterns expressed using abstract data types (in ML, a mode of use of the module system) in Section \ref{sec:syntax}.
-This  will allow  us to express  precisely the semantics that we seek, but syntactically,  the usual approach of using run-time string parsing whenever we introduce a pattern will leave something to be desired.
-To avoid run-time string parsing, we will be tempted to introduce derived syntactic forms specific to our library, thereby constructing a \emph{derived syntactic dialect}, i.e.
-a dialect specified by context-free elaboration to the existing language.
-Such derived syntactic dialects are not uncommon.
-For example Ur/Web is a  derived syntactic dialect of Ur (itself descended from ML) that builds in derived syntax for SQL queries, HTML elements and other datatypes common in web programming \cite{conf/popl/Chlipala15}.
-%In fact, nearly all major programming languages primitively build in derived syntax for constructs that can otherwise be expressed in the standard library, e.g.
+Many other types of data can similarly benefit from the availability of specialized derived syntax.
+For example, we will consider regular expression patterns expressed using abstract data types (as a mode of use of the ML module system) in Sec.
+\ref{sec:syntax}.
+%Though we will be able to express  precisely the semantics that we seek,   the usual approach of using dynamic string parsing to introduce a pattern will leave something to be desired syntactically.
+% An extension of Moscow ML provided syntax for monads encoded using the module system, inspired by the  syntax built in to Haskell.\todo{asked for reference, if not found will say that "Harper reports..."} %In fact, nearly all major programming languages primitively build in derived syntax for constructs that can otherwise be expressed in the standard library, e.g.
 derived list syntax in SML \cite{harper1997programming,mthm97-for-dart}.
-Tools like Camlp4 \cite{ocaml-manual} and Sugar* \cite{erdweg2011sugarj,erdweg2013framework} are designed to lower the engineering costs of constructing derived syntactic dialects, contributing to their continued proliferation (we will consider them further in Sec.
+Tools like Camlp4 \cite{ocaml-manual} and Sugar* \cite{erdweg2011sugarj,erdweg2013framework} are designed to lower the engineering costs of constructing derived syntactic dialects in situations like these, contributing to their continued proliferation.
+%(we will consider them further in Sec.
 \ref{sec:syntax}).
@@
 Record types themselves are perhaps  the simplest example -- they cannot be expressed by context-free elaboration to a language with only nullary and binary products.
-Various dialects of ML and Haskell have explored  ``record-like'' primitives that go further, supporting functional update and extension operators, width and depth coercions (sometimes implicit)%\cite{Cardelli:1984:SMI:1096.1098}
-, methods, prototypic dispatch and various other semantic ``embellishments'' that cannot be expressed by context-free elaboration to a language with standard record types (we will detail an example in Sec.
+Various dialects of ML and Haskell have explored  ``record-like'' primitives that go even further, supporting functional update and extension operators, width and depth coercions (sometimes implicit)%\cite{Cardelli:1984:SMI:1096.1098}
+, methods, prototypic dispatch and other semantic ``embellishments'' that cannot be expressed by context-free elaboration to a language with only standard record types (we will detail one  example in Sec.
 \ref{sec:metamodules}).
-The Ocaml dialect of ML builds in ``polymorphic variants'', ``open datatypes'' and  logic for typechecking operations that use format strings like $\mathtt{sprintf}$ \cite{ocaml-manual}.
+The OCaml dialect of ML primitively builds in ``polymorphic variants'', ``open datatypes'' and  logic for typechecking operations that use format strings like $\mathtt{sprintf}$ \cite{ocaml-manual}.
 ReactiveML builds in primitives for functional reactive programming \cite{mandel2005reactiveml}.
-ML5 builds in primitives for distributed programming \cite{Murphy:2007:TDP:1793574.1793585}.
-Manticore builds in parallel arrays \cite{conf/popl/FluetRRSX07}.
-Alice ML builds in futures and promises \cite{AliceLookingGlass}.
-MLj builds in the entire Java programming language, to support typesafe interoperation with Java libraries \cite{Benton:1999:IWW:317636.317791}.
-Tools like proof assistants and logical frameworks support specifying and reasoning metatheoretically about dialects like these, and tools like compiler generators and language workbenches lower the cost of implementing them, contributing to their  proliferation.
+ML5 builds in high-level primitives for distributed programming based on a modal lambda calculus \cite{Murphy:2007:TDP:1793574.1793585}.
+Manticore \cite{conf/popl/FluetRRSX07} and Alice ML  \cite{AliceLookingGlass} build in parallel programming primitives with a more elaborate type structure than is found in their respective core calculi.
+MLj builds in the type structure of the Java object system (motivated by a desire to interface safely and naturally with Java libraries) \cite{Benton:1999:IWW:317636.317791}.
+Other dialects do the same for other foreign languages, e.g.
+Furr and Foster describe a dialect of OCaml that builds in the type structure of C \cite{Furr:2005:CTS:1065010.1065019}.
+Tools like proof assistants and logical frameworks can be used to specify and reason metatheoretically about dialects like these, and tools like compiler generators lower the cost of implementing them, again contributing to their continued proliferation.
 
@@
 Unfortunately, taking this approach makes it substantially more difficult for clients to import high-level abstractions orthogonally.
-
-The reason why this proliferation of language dialects should be considered alarming is that it is, in an important sense,  anti-modular: a program written in one dialect cannot, in general, safely and idiomatically interface with libraries written in a different dialect.
+% test 
+The reason why this proliferation of language dialects should be considered alarming is that it is, in an important sense, anti-modular: a library written in one dialect cannot, in general, safely and idiomatically interface with a library written in another dialect.
 %At best, one can hope that the compilers for the decidedlytwo languages target a common intermediate language,
-To do so would fundamentally require combining the two dialects into a single language.
-Left unconstrained, this is an ill-posed problem, i.e.
-given two dialects that are specified by judgements of arbitrary form, it is not clear what it would even mean to combine them.
-Even when the dialects are specified using some formalism that does define what it means to combine two dialects, there is generally no guarantee that the composition will conserve important syntactic and semantic properties.
+As MLj demonstrates, addressing the interoperability problem requires somehow ``combining'' the  dialects into a single language.
+In the most general setting where the dialects in question can be specified by judgements of arbitrary form, this is not a well-defined notion.
+Even if we restrict our interest to dialects specified using formalisms that do operationalize some notion of dialect combination, there is generally no guarantee that the combined dialect will conserve syntactic and semantic properties that can be established about the dialects in isolation.
+%In other words, any putative ``combined language'' must formally be considered a  distinct system for which one must derive essentially all metatheorems of interest anew, guided only informally by those derived for the dialects individually.
 %There is no well-defined mechanism for constructing such a ``combined language'' in general.
-For example, consider two derived syntactic dialects, one building in derived syntax for JSON (a popular data interchange format), the other using a similar syntax for finite mappings of some other type.
-Though each is known to have an unambiguous concrete syntax in isolation, when their grammars are na\"ively  combined (e.g.
-by Camlp4), ambiguities arise.
-Any putative ``combined language'' must formally be considered a  distinct system for which one must derive all metatheorems of interest anew, guided only informally by those derived for the dialects individually.
+For example, consider two derived syntactic dialects specified using context-free grammars, one specifying derived syntax for finite mappings, the other specifying a similar syntax for ordered finite mappings.
+Though each can be shown to have an unambiguous concrete syntax in isolation, when their grammars are na\"ively  combined by, for example, Camlp4,  ambiguities  arise.
 % would generally attempt to combine their semantics, particularly when they are specified in very different ways.
@@
 %It it clear that dialects are better rhetorical devices than practical engineering artifacts. 
-Due to this paucity of modular reasoning principles, language dialects are not practical artifacts for software development ``in the large''.
+Due to this paucity of modular reasoning principles, language dialects are not practical for software development ``in the large''.
 %Large software projects and software ecosystems must pick a single language that does provide powerful modular reasoning principles and, to benefit from them, stay inside it.
 
-Dialects do, however, have a more indirect impact on large-scale software development in that the language designers that control comparatively popular languages, like Ocaml and Scala, can be convinced to incorporate  ideas from dialects  into backwards compatible language revisions.
+Dialect designers must ultimately take a less direct approach to have an impact on large-scale software development -- they must convince the designers in control of comparatively popular languages, like Ocaml and Scala, to include some suitable variant of the primitives they espouse into backwards compatible language revisions.
 %These decisions are increasingly influenced by community processes, e.g.
@@
 %Dialects thus serve the role of rhetorical vehicles for new ideas, rather than direct artifacts. 
-However, this has come at a cost: these languages have ballooned in size.
-This is not a scalable approach -- as suggested by the diversity of examples above, there is a ``long tail'' of potentially useful such language extensions (for derived syntax, our group has  gathered initial data speaking to this \cite{TSLs}).
+%Over time, accepting such extensions has caused these languages to balloon in size. 
+This \emph{ad hoc} approach is not sustainable, for two main reasons.
+First, as suggested by the diversity of examples given above, there are simply too  many potentially useful such primitives, and many of these are only relevant in relatively narrow application domains (for derived syntax, our group has  gathered initial data speaking to this \cite{TSLs}).
+Second, primitives introduced earlier in a language's lifespan end up monopolizing finite ``syntactic resources'', forcing subsequent primitives to use ever more esoteric forms.
+A related problem is that primitives cannot be removed or changed without breaking backwards compatibility.
 %rimitives that are only situationally useful, or that trigger aesthetic disagreements between different factions of the community, are still quite often left languishing in impractical ``toy'' dialects.
-Recalling the words of  Reynolds, which are as relevant today as they were nearly half a century ago \cite{Reynolds70}:%Because there is no data about how useful a construct is in practice until it is included in a language like this, decisions about which constructs to include are often informed by little more than intuition.
+Recalling the words of  Reynolds, which are as relevant today as they were almost half a century ago \cite{Reynolds70}:%Because there is no data about how useful a construct is in practice until it is included in a language like this, decisions about which constructs to include are often informed by little more than intuition.
 %This approach is antithetical to the ideal of a truly \emph{general-purpose language} described at the beginning of this section.
@@
 and generality in language design is a serious unsolved 
-problem.}\begin{flushright}--- John Reynolds (1970)\end{flushright}
+problem.}%\begin{flushright}--- John Reynolds (1970)\end{flushright}
 \end{quote}
@@
 \newpage %A measured approach to incorporating new primitives into a general-purpose programming language is  sensible because once a primitive is introduced, it becomes entrenched as-is and monopolizes ``syntactic resources'', as we will discuss below.
-This leaves the language design community with two possible paths forward if we wish to keep general-purpose languages small and free of \emph{ad hoc} primitives.
-One, exemplified (arguably) by SML, is to generally eschew the introduction of new primitives and settle on a set of primitives that sit at a ``sweet spot'' in the overall language design space, accepting that in some circumstances, this trades away expressive power or leads to  high syntactic cost.
+This leaves the subset of the  language design community interested in keeping general-purpose languages small and free of \emph{ad hoc} primitives with two possible paths forward.
+One, exemplified (arguably) by SML, is to simply eschew the introduction of specialized syntax and type structure and settle on the existing primitives, which can be said to sit at a ``sweet spot'' in the overall language design space (accepting that in some circumstances, this trades away expressive power or leads to  high syntactic cost).
 % that this implies.
-The other path forward is to continue on in the search for even more general language primitives, i.e.
-primitives that are so expressive that they permit us to degrade a broad class of existing primitives, including those found in dialects, to modularly composable library constructs, so that they can be evaluated on their individual merits by programmers and used together without the possibility of conflict.
+The other is to search for even more general primitives that reduce these seemingly \emph{ad hoc} primitives to modularly composable library constructs.
+% so that they can be evaluated on their individual merits by programmers and used together without the possibility of conflict.
 % This should render the construction of new dialects increasingly unnecessary.
-Encouragingly, primitives that have such a ``slimming'' effect on a language do occasionally arise.
+Encouragingly, primitives of this sort do occasionally arise.
 For example, a recent revision of Ocaml added support for  ``generalized algebraic data types'' (GADTs), based on research on guarded recursive datatype constructors \cite{XiCheChe03}.
-Using GADTs, Ocaml was able to move some of the machinery for typechecking operations that use format strings, like \texttt{sprintf}, out of the language and into the standard library (but some machinery remains built in).
-Our broad aim in the work being proposed is to introduce a new functional programming language called Verse\footnote{We distinguish Verse from Wyvern, which is the language referred to in prior publications about some of the work being proposed here, because Wyvern is a group effort evolving independently in some important ways.} that takes  more dramatic steps down this second path by allowing library providers to express new derived syntax and type structure of their own design, in a safe and modularly composable manner.
+Using GADTs, Ocaml was able to move some of the \emph{ad hoc} machinery for typechecking operations that use format strings, like \texttt{sprintf}, out of the language and into a library (however, syntactic machinery remains built in).
 
@@
 \section{Proposed Contributions}
-
-Verse features a module system taken directly from SML, but its core language is organized in a novel manner as a \emph{type-directed translation semantics} targeting a minimal typed lambda calculus called the Verse \emph{internal language} (IL).
-We formally introduce Verse's core semantics in Sec.
+Our broad aim in the work being proposed is to introduce a new functional programming language called Verse\footnote{We distinguish Verse from Wyvern, which is the language referred to in prior publications about some of the work being proposed here, because Wyvern is a group effort evolving independently in some important ways.} that takes more radical steps down the second path just described by giving library providers the ability to directly express derived syntax and type structure of a variety of designs in a safe and modularly composable manner.
+%This eliminates the needs to build in fewer \emph{ad hoc} constructs and dialects are less frequently necessary.
+%thereby reducing the need for language dialects and revisions.
+
+Verse features a module system taken directly from  SML.
+Unlike SML, the Verse core language is split into a \emph{typed external language} (EL) specified by {type-directed translation} to a minimal \emph{typed internal language} (IL).
+We formally introduce the semantics of the Verse core language in Sec.
 \ref{sec:verse}.
-In lieu of the typical \emph{external} (i.e.
-programmer-facing) derived syntax and  type structure, Verse introduces two novel primitive constructs:
+Our novel contributions will be two primitive constructs that eliminate the need for many others:
 \begin{itemize}
 \item \textbf{Typed syntax macros} (TSMs), introduced in Sec.
-\ref{sec:syntax}, subsume the need to primitively build in derived concrete syntax specific to library constructs, e.g.
+\ref{sec:syntax}, eliminate the need to primitively build in derived concrete syntax specific to library constructs, e.g.
 list syntax as in SML or XML syntax as in Scala and Ur/Web, by giving library providers static control over the parsing and elaboration of delimited segments of concrete syntax.
@@
 \item \textbf{Metamodules}, introduced in Sec.
-\ref{sec:metamodules}, subsume the need to primitively build in the type structure of constructs like records (and variants thereof),  labeled sums and other interesting constructs by giving library providers static hooks directly into the type-directed translation semantics.
-For example, a metamodule that implements the type structure of records would:
+\ref{sec:metamodules}, eliminate the need to primitively build in the type structure of constructs like records (and variants thereof),  labeled sums and other more esoteric constructs that we will consider by giving library providers static hooks directly into Verse's type-directed translation semantics.
+For example, a library provider can implement the type structure of records by defining a metamodule that:
 \begin{enumerate}
-\item Introduce a type constructor \lstinline{record} parameterized by finite mappings from labels to types, defining its semantics by translation to (for example) unary and binary products (which are built in to the internal language); and
-\item Introduce operators used to work with records, minimally record introduction and elimination, but perhaps also various functional update operators, defining the logic governing their typechecking and translation to the IL directly.
+\item Introduces a type constructor, \lstinline{record}, parameterized by finite mappings from labels to types and defines its semantics by translation to unary and binary products (which are built in to the internal language).
+\item Introduces operators used to work with records, minimally record introduction and elimination (but perhaps also various functional update operators), and directly defines the logic governing their typechecking and translation.
 \end{enumerate}
-We will see direct analogies between modules and metamodules as we continue.
+We will see direct analogies between modules and metamodules in Sec. \ref{sec:metamodules}.
 \end{itemize} 
-Both TSMs and metamodules can be understood as \emph{metaprogramming} primitives, because their static semantics involves the invocation of user-defined functions.% static code generation.
+Both TSMs and metamodules can be understood as \emph{metaprogramming} primitives, because their static semantics will involve the evaluation of user-defined functions that manipulate static representations of types and terms.% static code generation.
 %We will  also introduce a simple variant of each of these primitives that leverages Verse's support for local type inference to further reduce syntactic cost in certain common situations.
@@
 The key challenge in the design of these primitives will come in ensuring that they are metatheoretically well-behaved, given that they aim to decentralize control over aspects of the language's concrete syntax and type structure that, in other contemporary languages, are under the exclusive control of the language designer.
-If we are not careful, many of the problems discussed earlier as inherent to combining distinct language dialects could simply shift into the semantics of these primitives.\footnote{This is why languages  like Verse are often called \emph{extensible languages}, though this is somewhat of a misnomer.
+If we are not careful, many of the problems discussed earlier that arise when combining language dialects could simply shift into the semantics of these primitives.\footnote{This is why languages  like Verse are often called \emph{extensible languages}, though this is somewhat of a misnomer.
 The chief characteristic of an extensible language is that it \emph{doesn't} need to be extended in situations where other languages would need to be extended.
 We will avoid this somewhat confusing terminology.} Our main technical contributions will come in showing how to address these problems in a principled manner.
-In particular, syntactic conflicts will be made impossible by construction and  we will validate the code  statically generated by TSMs and metamodules to maintain a \emph{hygienic type discipline} and, most uniquely, powerful \emph{modular reasoning principles}.
-In other words, library providers will have the ability to reason about the constructs that they define in isolation, and clients will be able to use them safely in any combination, without the possibility of conflict.\footnote{We  assume that naming conflicts can be avoided extrinsically, e.g.
+In particular, syntactic conflicts will be impossible by construction and the semantics will validate code statically generated by TSMs and metamodules to maintain a strong \emph{hygienic type discipline} and, most uniquely, powerful \emph{modular reasoning principles}.
+In other words, library providers will have the ability to reason about the constructs that they have defined in isolation, and clients will be able to use them safely in any program context and in any combination, without the possibility of conflict.\footnote{This is not quite true -- simple naming conflicts can arise.
+We will tacitly assume that they are being avoided extrinsically, e.g.
 by using a URI-based naming scheme as in the Java ecosystem.}
@@
 
-Finally, we are not interested here in languages that feature full-spectrum dependent types, which blur the phase separation between compile-time and run-time, though we conjecture that the primitives we introduce could be introduced into languages like Gallina (the ``external language'' of the Coq proof assistant) with some modifications.
-Verse should be compared to languages that maintain a phase separation, like ML, Haskell and Scala.
+Finally, we will not be interested here in languages that feature full-spectrum dependent types, which blur the phase separation between compile-time and run-time, though we conjecture that the primitives we introduce could be introduced into languages like Gallina (the ``external language'' of the Coq proof assistant) with some reasonable modifications.
+Verse is designed for use in situations where ML, Haskell or Scala would be suitable today.
 %Our interest is in making sure that the standard library is no more privileged than any other library.
@@
 
-\section{Summary of Judgements}\label{sec:verse}
-To situate ourselves within a formal framework, let us begin with a brief summary of how Verse is organized and specified. %Verse consists of a \emph{module language} atop a \emph{core language}.
+\section{Overview of Judgements}\label{sec:verse}
+To situate ourselves within a formal framework, let us begin with a brief overview of how Verse is organized and specified. %Verse consists of a \emph{module language} atop a \emph{core language}.
 
@@
 
-The core language is organized much like the first stage of a type-directed compiler (e.g.
+The Verse core language will be the focus of our formal efforts.
+It is organized much like the first stage of a type-directed compiler (e.g.
 the TIL compiler for Standard ML \cite{tarditi+:til-OLD}), consisting of a user-facing \emph{typed external language} (EL) specified by type-directed translation to a minimal \emph{typed internal language} (IL).
-The main judgements in the specification of the EL take the following form  (omitting various contexts for now):\footnote{Note that we use the terms ``expression'' and ``type'' without qualification to refer to  external expressions and types, respectively.
+The main judgements in the specification of the EL take the following form  (omitting various contexts for now):\footnote{Note that we will use the terms ``expression'' and ``type'' without qualification to refer to  external expressions and types, respectively.
 }
@@
 
-Features like state, exceptions, simple concurrency primitives, scalars, arrays and others characteristic of a first-stage compiler intermediate language would also be included in the IL in practice, and in some cases this would affect the shape of the internal semantics, but we omit them for simplicity.
+Features like state, exceptions, simple concurrency primitives, scalars, arrays and others characteristic of a first-stage compiler intermediate language would also be included in the IL in practice, and in some cases this would affect the shape of the internal semantics, but we will omit them for simplicity.
 %Features like record types and labeled sum types, on the other hand, are not built in.
@@
 \section{Modularly Metaprogrammable Textual Syntax}\label{sec:syntax}
-Verse, like most major contemporary programming languages, specifies a textual concrete syntax.
-%We have chosen to specify a layout-sensitive textual concrete syntax (i.e.
+Verse, like most major contemporary programming languages, specifies a textual concrete syntax.\footnote{Although Wyvern specified a layout-sensitive concrete syntax, to avoid unnecessary distractions, we will specify a more conventional layout-insensitive concrete syntax for Verse.} %We have chosen to specify a layout-sensitive textual concrete syntax (i.e.
 newlines and indentation are not ignored).
@@
 %For constructs that have an obvious analog in ML, we will omit a detailed explanation.
-A language's concrete syntax serves as its programmer-facing user interface, so it is common practice to build in  derived forms (colloquially, \emph{syntactic sugar}) that capture common idioms more concisely or naturally.
+Because the purpose of concrete syntax is to serve as the language's programmer-facing user interface, it is common practice to build in  derived syntactic forms (colloquially, \emph{syntactic sugar}) that capture common idioms more concisely or naturally.
 % (i.e.
 considering cognitive dimensions \cite{green1996usability}).
-For example, derived list syntax is built in to most dialects of ML, so that instead of having to write \lstinline{Cons(1, Cons(2, Cons(3, Nil)))}, a programmer can equivalently write \lstinline{[1, 2, 3]}.
-Many languages go further, building in syntax associated with various other types of data, like vectors (SML/NJ), arrays (Ocaml), commands (Haskell), syntax trees (Scala), XML data (Scala, Ur/Web) and SQL queries (F\#, Ur/Web).
+For example, derived list syntax is built in to most functional languages, so that instead of having to write \lstinline{Cons(1, Cons(2, Cons(3, Nil)))}, the programmer can equivalently write \lstinline{[1, 2, 3]}.
+Many languages go further, building in derived syntax associated with various other types of data, like vectors (SML/NJ), arrays (Ocaml), monadic commands (Haskell), syntax trees (Scala), XML documents (Scala, Ur/Web) and SQL queries (F\#, Ur/Web).
 %This is a rather \emph{ad hoc} process.% discussed previously, the usual approach is to require that the language designer build in new derived syntactic forms.
@@
 
-Rather than privileging  particular library constructs with primitive syntactic support, Verse exposes primitives that allow library providers to introduce new derived syntax on their own.
+Verse takes a less \emph{ad hoc} approach -- rather than privileging  particular library constructs with primitive syntactic support, Verse exposes primitives that allow library providers to introduce new derived syntax of a variety of designs on their own.
 %Lists need no special consideration from the language specification.
-%The purpose of this section is to motivate and then introduce Verse's syntax extension mechanisms.
+%The purpose of this section is to  a and then introduce Verse's syntax extension mechanisms.
 %For forms with a clear analogy to a form in Standard ML, we will assume the  semantics are analagous without providing details.
-To specifically motivate our desire for this level of syntactic control, we begin in Sec.
-\ref{sec:examples} with a representative example: regular expression patterns expressed using abstract data types.
-We show how the usual approach of using strings to introduce patterns is not ideal.
+We will begin in Sec.
+\ref{sec:examples} by considering a representative example in detail: regular expression patterns expressed using abstract data types.
+We show how the usual approach of using strings to introduce patterns is less than ideal.
 We then survey existing alternatives in Sec.
@@
 \subsection{Motivating Example: Regular Expression Syntax}\label{sec:examples}
-Let us begin from the perspective of a regular expression library provider.
+Let us begin by taking the perspective of a regular expression library provider.
 Recall that regular expressions are a common way to capture patterns in strings \cite{Thompson:1968:PTR:363347.363387}.
 We will assume that the abstract syntax of {patterns}, $p$, over strings, $s$, is specified as below:\[p ::= \textbf{empty} ~|~ \textbf{str}(s) ~|~ \textbf{seq}(p; p) ~|~ \textbf{or}(p; p) ~|~ \textbf{star}(p) ~|~ \textbf{group}(p)\]
-The most direct way to express this abstract syntax is by defining a recursive sum type \cite{pfpl}.
-Verse supports these as case types, which are analagous to datatypes in ML (we will see how the type structure  of case types can in fact be expressed in libraries later):
+One way to express this abstract syntax is by defining a recursive sum type \cite{pfpl}.
+Verse supports these as case types, which are comparable to datatypes in ML (we plan to show how the type structure  of case types can in fact be expressed in libraries later):
 
@@
 However, there are some reasons not to expose this representation of patterns directly to clients.
-First, such patterns are usually identified up to their reduction to a normal form.
+First, regular expression patterns are usually identified up to their reduction to a normal form.
 For example, $\textbf{seq}(\textbf{empty}, p)$ has normal form $p$.
@@
 Indeed, there may be many ways to represent regular expression patterns, each with different performance trade-offs.
-For these reasons, a better approach is to use \emph{abstract data types}, which in Verse, as in ML, are a mode of use of the module system.
-In particular, we can define the following \emph{module signature}, where the type of patterns, \lstinline{t}, is held abstract.
-The client of any module \lstinline{P : PATTERN} can then identify patterns as terms of type \verb|P.t|.
+For these reasons, a better approach in Verse, as in ML, is to abstract over the choice of representation using  the module system.
+In particular, we can define the following \emph{module signature}, where the type of patterns, \lstinline{t}, is held abstract and functions are provided in lieu of datatype constructors and case analysis:
 %Notice that it exposes an interface otherwise  to the one available using a case type:
@@
 \begin{lstlisting}[deletekeywords={case},numbers=none]
-signature PATTERN = sig 
+signature PATTERN = sig {
   type t
@@
     'a)
-end
-\end{lstlisting}
-By holding the representation type of patterns abstract, the burden of proving that the case analysis function above cannot be used to distinguish patterns with the same normal form is localized to each module implementing this signature.
-The details are standard and not particularly interesting given our purposes, so we omit them.
+}
+\end{lstlisting}
+ A client of a module \lstinline{P} that matches \lstinline{PATTERN}, which we write \lstinline{P : PATTERN}, can express patterns as terms of type \verb|P.t|.
+By holding the representation type of patterns abstract, the burden of proving that the case analysis function cannot be used to distinguish patterns with the same normal form is localized to each module implementing this signature.
+The details are standard and not particularly relevant to our purposes, so we omit them here.
 
@@
 To express the concrete syntax of patterns, regular expression library providers usually provide a utility function that transforms strings to patterns.
-Because, as just mentioned, there  may be many implementations of the \lstinline{PATTERN} signature, the standard approach is to define a \emph{parameterized module} (a.k.a.
-\emph{functor} in SML) defining utility functions generically:
-
-\begin{lstlisting}[numbers=none]
-module PatternUtil(P : PATTERN) = mod 
+Because, as just mentioned, there  may be many implementations of the \lstinline{PATTERN} signature, the standard approach is to define a parameterized module (a.k.a.
+a \emph{functor} in SML) defining utility functions like this abstractly:
+
+\begin{lstlisting}[numbers=none]
+module PatternUtil(P : PATTERN) => mod {
   fun parse(s : string) : P.t = (* ... pattern parser here ... *)
-end 
+}
 \end{lstlisting}
@@
 \begin{lstlisting}[numbers=none]
-module PU = PatternUtil(P)
-let pattern = PU.parse
+let module PU = PatternUtil(P)
+let val pattern = PU.parse
 \end{lstlisting}
@@
 \begin{lstlisting}[numbers=none,mathescape=|]
-let ssn = pattern "SSTR\d\d\d-\d\d-\d\d\d\dESTR"
+let val ssn = pattern "SSTR\d\d\d-\d\d-\d\d\d\dESTR"
 \end{lstlisting}
 When compiling an expression like this, the programmer would see an error message like \verb|error: illegal escape character|.\footnote{This is the error message that \texttt{javac} produces.
-When compiling an analagous expression using SML of New Jersey (SML/NJ), we encounter the rather bizarre error message \texttt{Error: unclosed string}.} In a small lab study, we observed that this sort of error was common, and that solving the problem was nearly always initially challenging to even experienced programmers who hadn't used regular expressions recently \cite{Omar:2012:ACC:2337223.2337324}.
-The inelegant workaround is to  double backslashes:
-\begin{lstlisting}[numbers=none]
-let ssn = pattern "SSTR\\d\\d\\d-\\d\\d-\\d\\d\\d\\dESTR"
+When compiling an analagous expression using SML of New Jersey (SML/NJ), we encounter the rather bizarre error message \texttt{Error: unclosed string}.} In a small lab study, we observed that this sort of error was common, and that diagnosing the problem was nearly always initially challenging to even experienced programmers who had not used regular expressions recently \cite{Omar:2012:ACC:2337223.2337324}.
+One inelegant workaround is to  double backslashes:
+\begin{lstlisting}[numbers=none]
+let val ssn = pattern "SSTR\\d\\d\\d-\\d\\d-\\d\\d\\d\\dESTR"
 \end{lstlisting}
@@
 \begin{lstlisting}[numbers=none]
-let ssn = pattern {rx|SSTR\d\d\d-\d\d-\d\d\d\dESTR|rx}
-\end{lstlisting}
-
-\item The next problem is that pattern parsing does not occur until the pattern is evaluated at run-time.
+let val ssn = pattern {rx|SSTR\d\d\d-\d\d-\d\d\d\dESTR|rx}
+\end{lstlisting}
+
+\item The next problem is that pattern parsing does not occur until the pattern is evaluated.
 For example, the following malformed pattern will only trigger an error when this expression is evaluated during the full moon: %Achieving this goal is an explicit goal of this proposal, so we are obviously not happy with this.
@@
 \end{lstlisting}
-Such issues can sometimes be found via testing, but empirical data gathered from open source projects suggests that there remain many malformed regular expression patterns that are not detected by a project's test suite ``in the wild'' \cite{spishak2012type}.
+Such problems can sometimes be discovered via testing, but empirical data gathered from large open source projects suggests that there remain many malformed regular expression patterns that are not detected by a project's test suite ``in the wild'' \cite{spishak2012type}.
 
 We might instead attempt to treat the well-formedness of patterns constructed from strings as a static verification condition.
-Statically proving that this condition holds throughout a program is wildly difficult to automate in general, because it involves reasoning about arbitrary dynamic behavior.
-In the example above, we must know that the variable \lstinline{pattern} binds the function \lstinline{PU.pattern}.
+Statically proving that this condition holds throughout a program is wildly difficult to automate, because it involves reasoning equationally about arbitrary dynamic behavior.
+In the example above, we must know that the variable \lstinline{pattern} is equal to the function \lstinline{PU.parse}.
 If the argument had not been written literally but rather computed, e.g.
-if we had written \lstinline{"SSTR(GESTR" ^ "SSTRCESTR"} where \lstinline{^} is the string concatenation function applied in the infix style, we would need to be able to establish that this expression is equivalent to the string above.
+if we had written \lstinline{"SSTR(GESTR" ^ "SSTRCESTR"} where \lstinline{^} is the string concatenation function applied in infix style, we would need to be able to establish that this expression is equivalent to the string above.
 For patterns that are dynamically constructed based on input to the program (further discussed below), the problem is harder still.
 
-Parsing patterns at run-time also incurs a performance penalty.
-To avoid incurring it every time the pattern is encountered (e.g.
-for each item in a very large dataset), one must be careful to stage the computation appropriately, or use an appropriately tuned caching strategy.
+\item Parsing patterns dynamically also incurs a performance cost. If the programmer is not careful, they might  incur this cost many times. For example:
+\begin{lstlisting}[numbers=none]
+map exmpl_list (fn x => rx_match (pattern "SSTRA|T|G|CESTR") x)
+\end{lstlisting}
+To avoid incurring the parsing cost for each element of \lstinline{exmpl_list}, the programmer or compiler must move the parsing step out of the transformation or ensure that an appropriately tuned caching strategy is being used.
+%The author's experience in bioinformatics suggests that often, none of these strategies is used.
 %This too is difficult if a portion of the pattern is dynamically generated.
@@
 
-\item The final problem is that using strings to construct patterns encourages programmers to use  operations over strings, like concatenation, to construct patterns derived from user input that happens also to be of type \lstinline{string}.
+\item The final problem is that using strings to construct patterns encourages programmers to use  operations over strings, like concatenation, to construct patterns derived from data that happens also to be of type \lstinline{string}.
 This can lead to subtle but serious bugs.
@@
 
-	\subsection{Existing Approaches}\label{sec:syntax-existing} The difficulties described above are common whenever strings are parsed dynamically to address the issue of syntactic cost (a static issue).
+	\subsection{Existing Approaches}\label{sec:syntax-existing} The difficulties described above are not unique to regular expression patterns.
+Whenever a library  encourages the use of dynamic string parsing solely to address the issue of syntactic cost (a fundamentally static issue), such problems arise.
 %Strings are, simply put, not ideally suited for this task.
-	This has motivated a substantial quantity of research on reducing this use of run-time string parsing \cite{Bravenboer:2007:PIA:1289971.1289975}.
-The existing approaches can be classified in  one of two ways: as being based on \emph{syntax extension} or \emph{static term rewriting}.
-
-\subsubsection{Syntax Extension}\label{sec:syntax-extension}
-One approach we might consider is to directly extend the syntax of our language with new derived syntactic forms.% for patterns.
+	This has motivated much research on reducing the need for dynamic string parsing \cite{Bravenboer:2007:PIA:1289971.1289975}.
+Existing approaches can be broadly classified as being based on either \emph{direct syntax extension} or \emph{static term rewriting}, described in Secs.
+\ref{sec:syntax-extension} and \ref{sec:term-rewriting}, respectively, below.% We summarize the most relevant work below.
+
+\subsubsection{Direct Syntax Extension}\label{sec:syntax-extension}
+One approach we might consider is to use a system that allows us to directly extend the syntax of our language with new derived syntactic forms for regular expression patterns.% for patterns.
 
@@
 For example, sequences of characters  can only be parsed as identifiers using these systems, rather than as characters in a pattern.
+
 Syntax extension systems based on context-free grammars like  Sugar* \cite{erdweg2013framework}, Camlp4 \cite{ocaml-manual} and many others are more expressive, and would allow us to directly introduce pattern syntax into our core language's grammar, perhaps like this:
 \begin{lstlisting}[numbers=none]
-let ssn = /\d\d\d-\d\d-\d\d\d\d/
-\end{lstlisting}
-
-However, this is a rather perilous approach because none of these systems guarantee freedom from \emph{syntactic conflicts}, i.e.
-as the Coq manual states: ``mixing different symbolic notations in [the] same text may cause serious parsing ambiguity''.
-So if another library provider used similar syntax for a different implementation of regular expressions, or for some other unrelated abstraction, then a client could not simultaneously use both libraries in the same scope.
-Properly considered, every combination of extensions introduced using one of these mechanisms creates a \emph{de facto} derived syntactic dialect of our language.
+let val ssn = /\d\d\d-\d\d-\d\d\d\d/
+\end{lstlisting}
+
+Unfortunately, this approach is rather perilous because none of the systems mentioned thusfar can guarantee that \emph{syntactic conflicts} between such extensions will not arise.
+As stated directly in the  Coq manual: ``mixing different symbolic notations in [the] same text may cause serious parsing ambiguity''.
+Consequently, if another library provider used similar syntax for a different implementation or variant of regular expressions, or for some other unrelated abstraction, then a client could not simultaneously use both libraries in the same scope.
+Properly considered, every combination of extensions introduced using these mechanisms creates a \emph{de facto} derived syntactic dialect of our language.
 % Resolving such parsing amibiguities is left to each client of the library.
@@
 
-Putting this aside, we must now decide how our newly introduced forms elaborate to forms in our base grammar.
+Putting this aside, we must also decide how our newly introduced forms elaborate to forms in our base grammar.
 This is tricky because we have defined a modular encoding of patterns -- which particular module should the elaboration use?
@@
 \begin{lstlisting}[numbers=none]
-let ssn = edu_cmu_verse_rx_pattern P /\d\d\d-\d\d-\d\d\d\d/
-\end{lstlisting}
-For patterns constructed compositionally, we can define \emph{splicing} syntax for both strings and other patterns, e.g. as shown below:
+let val ssn = edu_cmu_verse_rx_pattern P /\d\d\d-\d\d-\d\d\d\d/
+\end{lstlisting}
+For patterns constructed compositionally, we can define \emph{splicing} syntax for both strings and other patterns, e.g. rewriting \lstinline{example_bad} as shown below:
 \begin{lstlisting}[numbers=none,escapechar=|]
-fun example_syn(name : string) => 
+fun example_synexts(name : string) => 
   edu_cmu_verse_rx_pattern P /@name: %ssn/
@@
 
-\subsubsection{Static Term Rewriting}
+\subsubsection{Static Term Rewriting}\label{sec:term-rewriting}
 An alternative approach is to leave the concrete syntax of the language fixed, but repurpose it for novel ends using a \emph{local term-rewriting system}.
@@
 \begin{lstlisting}[numbers=none]
-let ssn = pattern P {rx|SSTR\d\d\d-\d\d-\d\d\d\dESTR|rx}
+let val ssn = pattern P {rx|SSTR\d\d\d-\d\d-\d\d\d\dESTR|rx}
 \end{lstlisting}
@@
 \subsection{Contributions}\label{sec:syntax-contributions}
-We propose a new primitive, the \textbf{typed syntax macro} (TSM), that combines the flexibility of syntax extensions with the reasoning guarantees of statically-typed macros, and show how to integrate it into a language with an ML-style module system, addressing all of the brought up thusfar.
-We then introduce \textbf{type-specific languages} (TSLs), which leverage local type inference to control TSM dispatch, further decreasing syntactic cost (which is, of course, the entire purpose of this exercise).
-The result is that derived syntax defined by library providers using TSMs and TSLs  approaches the cost of derived syntax built in primitively by a language designer (perhaps using a na\"ive tool like Camlp4).
-
-\subsubsection{Typed Syntax Macros (TSMs)}
+We propose a new primitive -- the \textbf{typed syntax macro} (TSM) -- that combines the flexibility of syntax extensions with the reasoning guarantees of statically-typed macros.
+This will address all of the problems brought up thusfar.
+We will introduce TSMs in Sec.
+\ref{sec:tsms} below.
+
+We then go further in Sec.
+\ref{sec:tsls}, proposing \textbf{type-specific languages} (TSLs).
+TSLs are TSMs associated directly with a type when it is generated.
+Verse then leverages local type inference to implicitly control TSL dispatch, further decreasing syntactic cost (which is, of course, the entire purpose of this exercise).
+
+The result is that derived syntax defined by library providers using TSMs and TSLs provides client with a syntactic cost cost comparable to that of derived syntax built in primitively by a language designer or  using a na\"ive tool like Camlp4, obviating the need for these approaches.
+
+\subsubsection{Typed Syntax Macros (TSMs)}\label{sec:tsms}
 %A typed syntax macro is invoked by applying it to a \emph{delimited form}, which can contain  arbitrary syntax in its \emph{body}.  
@@
 \end{lstlisting}
-Here, a parameterized TSM, \lstinline{pattern}, is applied to a module parameter, \lstinline{P}, and a {delimited form}, \lstinline{/SURLA|T|G|CEURL/}.
-The TSM statically parses the body of the provided delimited form, i.e.
-the characters between the delimiters (shown here in blue), and computes an elaboration, i.e.
+Here, a \emph{parameterized TSM}, \lstinline{pattern}, is being applied first to a module parameter, \lstinline{P}, then to a {delimited form}, \lstinline{/SURLA|T|G|CEURL/} (a number of alternative delimiters are also provided by Verse's concrete syntax and could equivalently be used).
+The TSM statically parses the \emph{body} of the provided delimited form, i.e.
+the characters between the delimiters (shown here in blue), and computes an \emph{elaboration}, i.e.
 another external expression.
-In this case, it is as if we wrote the following directly:
+In this case, the term above elaborates to the following term (written concretely):
 
@@
 %The original expression, above, is statically rewritten to this expression.
-The definition of \lstinline{pattern} is shown below:
-\begin{lstlisting}[numbers=none]
-syntax pattern(P : PATTERN) at P.t => 
+The definition of the parameterized TSM \lstinline{pattern} has the following basic form:
+\begin{lstlisting}[numbers=none]
+syntax pattern(P : PATTERN) at P.t {
   fn (ps : ParseStream) => (* pattern parser here *)
-\end{lstlisting}
-Note that identifying the module parameter as \lstinline{P} here is unimportant, i.e.
+}
+\end{lstlisting}
+Note that identifying the module parameter as \lstinline{P} here is not important, i.e.
 the TSM can be used with \emph{any} module matching signature \lstinline{PATTERN}.
-The module parameter is used in the type annotation \lstinline{at P.t}, which specifies that all elaborations successfully generated by applying this TSM to a module parameter \lstinline{P} will be of type \lstinline{P.t}.
-Elaborations are generated by the static action of the parse function defined in the indented block on the second line.
-The parse function must be of type \lstinline{ParseStream -> Exp}.
-Both types are defined in the Verse \emph{prelude}, which is a set of definitions available ambiently.
-The type \lstinline{ParseStream} gives the function access to the {body} of the delimited form (in blue above) and the type \lstinline{Exp} is a case type that encodes the abstract syntax of external expressions.
-
-For concision in analytic positions, i.e. those where an expected type is known due to, for example, a type ascription, the module parameter \lstinline{P} can be inferred:
-\begin{lstlisting}[numbers=none]
-let N : P.t = pattern /SURLA|T|G|CEURL/
-\end{lstlisting}
-
-TSMs can be partially applied and abbreviated using a let-binding style mechanism:
-\begin{lstlisting}[numbers=none]
+The module parameter is used in the type annotation \lstinline{at P.t}, which specifies that all elaborations successfully generated by applying this TSM to a module parameter \lstinline{P} and a delimited form will be of type \lstinline{P.t}.
+Elaborations are generated by the static action of the parse function defined on the second line.
+This parse function must be  a static value of type \lstinline{ParseStream -> Exp}.
+Both of these types are defined in the Verse \emph{prelude}, which is a set of definitions available ambiently.
+The type \lstinline{ParseStream} will give the function access to the {body} of the delimited form (in blue above) and the type \lstinline{Exp} is a case type that encodes the abstract syntax of external expressions.
+In other words, this function parses the body of the delimited form to generate an encoding of the elaboration.
+
+The parse function can identify portions of the body of the delimited form that should be treated as spliced expressions, allowing us to support splicing syntax as described in Sec.
+\ref{sec:syntax-extension}:
+\begin{lstlisting}[numbers=none]
+(* TSMs can be partially applied and abbreviated *)
 let syntax pat = pattern P
-let ssn = pat /SURL\d\d\d-\d\d-\d\d\d\dEURL/
-\end{lstlisting}
-
-TSMs can identify portions of the body of the delimited form that should be treated as spliced expressions, allowing us to support splicing syntax as described in Sec. \ref{sec:syntax-extension}:
-\begin{lstlisting}[numbers=none]
+let val ssn = pat /SURL\d\d\d-\d\d-\d\d\d\dEURL/
 fun example_tsm(name: string) => 
@@
 \end{lstlisting}
-A hygiene mechanism ensures that only portions of the generated elaboration derived from such spliced expressions can refer to the variables at the use site, preventing inadvertent variable capture by the elaboration.
-Put another way, the  elaboration logic must be valid in any context.
+%For concision in analytic positions, i.e. those where an expected type is known due to, for example, a type ascription, the module parameter \lstinline{P} can be inferred:
+%\begin{lstlisting}[numbers=none]
+%let N : P.t = pattern /SURLA|T|G|CEURL/
+%\end{lstlisting}
+
+A hygiene mechanism ensures that only portions of the generated elaboration derived from such spliced expressions can refer to the variables at the use site, preventing inadvertent variable apture by the elaboration.
+Put another way, the  elaboration logic must be valid in any typing context.
 
@@
 \item The first premise can be pronounced ``TSM expression $s$ is a TSM at type $\sigma$ under $\Delta$ with parser $\iota_\text{parser}$''.
-This judgement would handle the mechanisms of module parameter application, but we must elide the details here (a complete specification of this judgement represents the main avenue for remaining work).
+This judgement would handle the mechanisms of module parameter application, but we must elide the details here (a complete specification of this judgement represents the main piece of remaining work).
 % Application of a parameterized TSM to its parameters is handled by this judgement (not shown).
 \item The second premise creates a parse stream, $\iota_\text{ps}$, from the body of the delimited form.
-\item The third premise applies the parser to the parse stream to generate an encoding of the elaboration, $\iota_\text{elab}$.
+\item The third premise applies the parser to the parse stream to generate an encoding of the elaboration, $\iota_\text{elab}$. This is where the metaprogramming occurs.
 \item The fourth premise decodes $\iota_\text{elab}$, producing the elaboration $e_\text{elab}$.
@@
 
-Variables in the outer contexts cannot be used directly by the elaboration, e.g. they are ignored by the (syn-var) rule:
-\begin{mathpar}
-\inferrule[syn-var]{ }{
-    \Delta_\text{out}; \Gamma_\text{out}; \Delta; \Gamma, x : \sigma \vdash x \Rightarrow \sigma \leadsto x
-}
-\end{mathpar}
 The outer contexts are switched back in only when encountering a spliced expression, which is marked:
@@
 \end{lstlisting}
+
+Variables in the outer contexts cannot be used directly by the elaboration, e.g. they are ignored by the (syn-var) rule:
+\begin{mathpar}
+\inferrule[syn-var]{ }{
+    \Delta_\text{out}; \Gamma_\text{out}; \Delta; \Gamma, x : \sigma \vdash x \Rightarrow \sigma \leadsto x
+}
+\end{mathpar}
+
 \end{enumerate}
-
 %We plan to give a more complete account of all of these judgements in the dissertation.
-\subsubsection{Type-Specific Languages (TSLs)}
-To further lower the syntactic cost of using TSMs, Verse also  supports \emph{type-specific languages} (TSLs), which allow library providers to associate a TSM directly with a declared type.
+\subsubsection{Type-Specific Languages (TSLs)}\label{sec:tsls}
+To further lower the syntactic cost of using TSMs, Verse also  supports \emph{type-specific languages} (TSLs), which allow library providers to associate a TSM directly with a generated type.
 For example, a module \lstinline{P} can associate \lstinline{pattern} with \lstinline{P.t} as follows:
 \begin{lstlisting}[numbers=none]
-module P : PATTERN with syntax pattern(P) at t = mod 
+module P : PATTERN = mod {
   type t = (* ... *)
   (* ... *)
-end 
+} with syntax pattern at t
 \end{lstlisting}
@@
 \begin{lstlisting}[numbers=none]
-fun example_tsl(name : string) : P.t
+fun example_tsl(name : string) : P.t => 
   /SURL@EURLnameSURL: %EURLssn/
@@
   Nil
-  Cons of 'a * list('a)
-} with syntax
+| Cons of 'a * list('a)
+} with syntax {
   fn (body : ParseStream) => (* ... comma-delimited spliced exps ... *)
+}
 \end{lstlisting}
@@
 \begin{lstlisting}[numbers=none]
-let x : list(P.t) = [/SURL\dEURL/SHTML, EHTML/SURL\d\dEURL/SHTML, EHTML/SURL\d\d\dEURL/]
+let val x : list(P.t) = [/SURL\dEURL/SHTML, EHTML/SURL\d\dEURL/SHTML, EHTML/SURL\d\d\dEURL/]
 \end{lstlisting}
 From the client's perspective, it is essentially as if the language had built in derived syntax for lists and regular expression patterns directly.
-%The only constraint is that this syntax must be used in an analytic position, which we argue is actually better for code compren when encountering unfamiliar syntax.
+However, we did not need to build in this syntax primitively.%The only constraint is that this syntax must be used in an analytic position, which we argue is actually better for code compren when encountering unfamiliar syntax.
 \subsection{Timeline}\label{sec:syntax-timeline}
 We have described and given a more detailed formal specification of TSMs in a recently published paper \cite{sac15}, and TSLs in a paper last year \cite{TSLs}, both in the context of the Wyvern language.
-The mechanics of elaboration generation, encoding and decoding and hygienic elaboration validation have all been satisfyingly specified in this context.
-
-Wyvern does not specify an ML-style module system (type declarations provide a form of type generativity), so in the context of Verse, there are some changes that need to be made in the formalization to handle module parameters, as mentioned earlier.
-We plan to complete this in the course of writing the dissertation, and do not anticipate it requiring more than 1-2 weeks of effort over the summer.
+The mechanics of elaboration encoding/decoding and hygienic elaboration validation have all been satisfyingly specified in this context.
+
+Wyvern does not specify an ML-style module system (type declarations provide the only form of type generativity), so in the context of Verse, there are some extensions that need to be defined in the formalization to handle module parameters, as mentioned earlier.
+We plan to complete this in the course of writing the dissertation, and do not anticipate it requiring more than 2-3 weeks of effort over Summer 2015.
+We emphasize that we do not plan to specify the full module system in detail; we will assume a standard specification of the module system and specify only those portions that interface with the novel aspects of the core language.
 
@@
 Let us now turn our attention to the type structure of the Verse external language (EL).
-In contrast to other languages, which build in constructs like case types (shown briefly at the beginning of Sec.
-\ref{sec:examples}), tuples, records, objects and so on primitively, in Verse these can all be expressed as modes of use of \emph{metamodules} (which are distinct from, though conceptually related to, {modules}, as we will see).
+In contrast to other languages, which build in constructs like case types (shown briefly in use in the previous section), tuples, records, objects and so on primitively, in Verse these can all be expressed as modes of use of \emph{metamodules} (which are distinct from, though conceptually related to, {modules}, as we will see).
 Only polymorphic function types are built primitively into the Verse EL.
 
-This section is organized much like Section \ref{sec:syntax}.
-We begin with examples of type structure that we wish to express in Section \ref{sec:metamodules-example}, then discuss how existing approaches are insufficient in Section \ref{sec:metamodules-related}.
-Next, we show how metamodules (together with TSMs and TSLs) serve to address these problems in Section \ref{sec:metamodules-solution} and conclude with a timeline in Section \ref{sec:metamodules-timeline}.
+This section is organized much like Sec.
+\ref{sec:syntax}.
+We begin with examples of type structure that we wish to express in Sec.
+\ref{sec:metamodules-example}, then discuss how existing approaches are insufficient in Sec.
+\ref{sec:metamodules-related}.
+Next, we outline how metamodules (together with TSMs and TSLs) serve to address these problems in Sec.
+\ref{sec:metamodules-solution} and conclude with a timeline in Sec.
+\ref{sec:metamodules-timeline}.
 
 \subsection{Motivating Example: Labeled Tuples and Regular Strings}\label{sec:metamodules-example}
-As a simple introductory example, let us consider a type classifying conference papers:
+As a simple introductory example, let us define a type classifying conference papers:
 \begin{lstlisting}[numbers=none]
@@
 \begin{lstlisting}[numbers=none]
-let exmpl : Paper = {"SUSAn Example PaperEUS"SCSS, ECSS"SUSEXMPL 2015EUS"}
+let val exmpl : Paper = {"SUSAn Example PaperEUS"SCSS, ECSS"SUSEXMPL 2015EUS"}
 \end{lstlisting}
@@
 \begin{lstlisting}[numbers=none]
-let exmpl : Paper = {SCSSconf => ECSS"SUSEXMPL 2015EUS"SCSS, title => ECSS"SUSAn Example PaperEUS"SCSS}
+let val exmpl : Paper = {SCSSconf=>ECSS"SUSEXMPL 2015EUS"SCSS, title=>ECSS"SUSAn Example PaperEUS"SCSS}
 \end{lstlisting}
@@
 \begin{lstlisting}[numbers=none]
-let exmpl_conf = # <SNATconfENAT> exmpl 
+let val exmpl_conf = # <SNATconfENAT> exmpl 
 \end{lstlisting}
@@
 \begin{lstlisting}[numbers=none]
-let exmpl_conf_name = #group 0 exmpl_conf
+let val exmpl_conf_name = #group 0 exmpl_conf
 \end{lstlisting}
@@
 \begin{lstlisting}[numbers=none]
-let a : ltuple {SPCTauthors : EPCTlist(rstring /SURL.+EURL/)} = {["SUSHarry Q. BovikEUS"]}
-let exmpl_paper_final = ltuple+ exmpl_paper a
+let val a : ltuple {SPCTauthors : EPCTlist(rstring /SURL.+EURL/)} = {["SUSHarry Q. BovikEUS"]}
+let val exmpl_paper_final = ltuple+ exmpl_paper a
 \end{lstlisting}
@@
 \begin{lstlisting}[numbers=none]
-let exmpl_paper_anon = ltuple- <SNATauthorsENAT> exmpl_paper_final
+let val exmpl_paper_anon = ltuple- <SNATauthorsENAT> exmpl_paper_final
 \end{lstlisting}
@@
 \subsection{Existing Approaches}\label{sec:metamodules-related}
+Before considering labeled tuples and regular strings as a mode of use of metamodules, let us consider some alternative approaches.
+
 \subsubsection{Labeled Tuple Types}
@@
 For our example above:
-For example, we might define the following signature for our example type \lstinline{Paper} from earlier:
+For our example type \lstinline{Paper} from above, we start by defining the following signature:
 \begin{lstlisting}
-signature PAPER
+signature PAPER = sig {
   type t
@@
   val prj_conf : t -> conf_t
-\end{lstlisting}
-We've simply taken all possible valid introductory and projection operators and expressed them as functions, moving label parameters into identifiers.
-Even with our minimal EL, we could  implement this signature using a Church-style encoding (the details are standard and omitted).
-Alternatively, if we only slightly relax  our insistence on minimalism and add binary products to the EL, we could use these as well:
+}
+\end{lstlisting}
+We've simply taken all possible valid introductory and projection operators and called for their expression as  functions, moving label parameters into identifiers.
+Even with our minimal EL, we could  implement the semantics of these operators against this signature using a Church-style encoding (the details are standard and omitted).
+Alternatively, if we slightly relax  our insistence on minimalism and add binary products to the EL, we could use them to implement these operators as well:
 \begin{lstlisting}
- module Paper : PAPER
+ module Paper : PAPER = mod {
    type t = title_t * conf_t
-   fun intro_tpl title conf = (title, conf)
-   fun intro_title_conf title conf = (title, conf)
-   fun intro_conf_title conf title = (title, conf)
-   fun prj_title (title, conf) = title 
-   fun prj_conf (title, conf) = conf
-
- type Paper = Paper.t
+   fun intro title conf => (title, conf)
+   fun intro_title_conf title conf => (title, conf)
+   fun intro_conf_title conf title => (title, conf)
+   fun prj_title (title, conf) => title 
+   fun prj_conf (title, conf) => conf
+ }
  \end{lstlisting}
@@
 
-There are several basic problems with this approach.
-First,  each labeled tuple type that a library provider wishes to encode in this way requires a volume of boilerplate code factorial in the number of components it has (because we must consider all permutations of label orderings as a distinct introductory function).
-It is also not satisfying syntactically from the client's perspective.
-To avoid these problems and recover the syntax used previously, we could define a TSM for every such signature, though this too involves much boilerplate.
-
-More fundamentally, we could only express the introductory and projection operators.
-We were not able to enumerate all possible operators that arise from the opcon \lstinline{ltuple-} as functions, because to do so, we would need to construct an encoding of every labeled tuple type that might arise from dropping one or more components.
-This would require an exponential volume of boilerplate code.
-Finally, there is simply no way to define \lstinline{ltuple+} as a finite collection of functions -- there are infinitely many extensions of any particular labeled product type.
+However, there are several problems with this basic approach.
+First, not every module matching the signature \lstinline{PAPER} correctly implements the semantics of the operators we seek to express, so each such module would itself need to be verified.
+For example, we need to verify the universal condition that for all \lstinline{e : Paper.t} and \lstinline{f : Paper.t -> T}, we have that \lstinline{f(e)} is equivalent to \lstinline{f(Paper.intro (Paper.prj_title e) (Paper.prj_conf e))}.
+The volume of boilerplate code that needs to be generated and verified manually is factorial in the number of components of the labeled tuple type we are expressing (because we must encode each permutation of label orderings with a distinct introductory function).% Not all modules satisfying the signature above correctly implement the semantics, so this boilerplate must be checked for correctness in some way.
+
+Another issue is that we can only reasonably express the introductory and projection operators by enumerating them in this way.
+To enumerate all possible operators that arise from the opcon \lstinline{ltuple-} as functions would require constructing an encoding of every labeled tuple type that might arise from dropping one or more components.
+This would add an  exponential factor to our volume of boilerplate code.
+Finally, there is simply no way to define \lstinline{ltuple+} as a finite collection of functions because there are infinitely many extensions of any labeled tuple type.
 % of this operator would need to be implemented by clients manually, defeating the purpose of the operator.
@@
 
-The simplest approach we might consider taking is to directly map each labeled tuple type to a record type defining the same component mapping:
+The simplest approach we might consider taking is to directly map each labeled tuple type to a record type having the same component mapping:
 \begin{lstlisting}[numbers=none]
@@
 }\end{lstlisting}
-Unlike labeled tuple types, record types are identified only up to component reordering.
-In other words, this mapping does not preserve certain type disequalities between labeled tuple types.
-Lacking positional information, it is not possible to omit component labels and rely on positional information when introducing a value of this type directly.%, because this information is not maintained by the type system.
-
-To work around this limitation if we do not otherwise care about preserving these type disequalities, we would need to define two conversion functions involving tuples:
-\begin{lstlisting}[numbers=none]
-fun Paper_of_tpl (title : title_t, conf : conf_t)
+Unlike labeled tuple types, record types are identified only up to component reordering, so this mapping does not preserve certain type disequalities between labeled tuple types.
+This means it is not  possible to allow clients to omit component labels and rely on the positional information available in the type when introducing a value of this type.%, because this information is not maintained by the type system.
+
+To work around this limitation if we do not otherwise care about preserving these type disequalities, we could define two conversion functions involving unlabeled tuples:
+\begin{lstlisting}[numbers=none]
+fun Paper_of_tpl (title : title_t, conf : conf_t) => 
   {title => title, conf => conf}
-fun tpl_of_Paper {title => title, conf => conf} = (title, conf)
-\end{lstlisting}
-For providers, this has syntactic cost linear in the number of components.
-For clients, we must explicitly invoke these functions, which incurs both a syntactic and run-time cost.
-We could define a TSM for every such type to reduce this cost to clients, though this is an additional cost to providers.%This also means that we cannot omit the labels and rely on positional information when introducing a value of a record type in an analytic position, as we did above with type \lstinline{Paper}.
+fun tpl_of_Paper {title => title, conf => conf} =>
+  (title, conf)
+\end{lstlisting}
+For clients to be able to rely on this approach, we must extrinsically verify that the library provider implemented these functions correctly.
+Clients must also bear the syntactic and run-time costs of explicitly invoking these functions.
+We could define a TSL for every such type to reduce this cost to clients, though this is an additional cost to providers.
+%This also means that we cannot omit the labels and rely on positional information when introducing a value of a record type in an analytic position, as we did above with type \lstinline{Paper}.
 %If we instead use an unlabeled product type (which in ML is actually a record type with numeric labels, which are omitted syntactically), we lose the ability to use labels in the elimination form.
@@
 \begin{lstlisting}[numbers=none]
-fun Paper_of_tpl (title : title_t, conf : conf_t) = {
+fun Paper_of_tpl (title : title_t, conf : conf_t) => {
   __0_title => title,
@@
 }
-fun tpl_of_Paper {__0_title => title, __1_conf => conf} = (title, conf)
-fun Paper_of_rcd {title => title, conf => conf} = {
+fun tpl_of_Paper {__0_title => title, __1_conf => conf} => 
+  (title, conf)
+fun Paper_of_rcd {title => title, conf => conf} => {
   __0_title => title,
@@
 \end{lstlisting}
-Clients again bear the cost of applying these conversion functions ahead of every operation and providers again bear the burden of defining this boilerplate (which is again linear in the number of components, albeit with a constant factor that is again not easy to dismiss).
+Clients again bear the cost of applying these conversion functions ahead of every operation and providers again bear the burden of defining this boilerplate code (which again has volume linear in the number of components, albeit with a constant factor that is again not easy to dismiss).
+And once again, the provider must verify that these functions were implemented correctly.
 %We are interested in all such pragmatic issues, so this is not an entirely satisfying solution.
@@
 
-As with our modular encoding above, neither of these workarounds provides us with any way to uniformly express the more interesting operations, like  concatenation, that we discussed earlier.
+Finally, note that neither of these workarounds provides us with any way to uniformly express the more interesting operations, like  concatenation or dropping a component, that we discussed earlier.
 
@@
 \begin{lstlisting}[numbers=none]
-let conf : string (* ~ rstring /([A-Z]+) (\d\d\d\d)/ *) = "EXMPL 2015"
+let val conf : string (* ~ rstring /([A-Z]+) (\d\d\d\d)/ *) = "EXMPL 2015"
 \end{lstlisting}
@@
 \begin{lstlisting}[numbers=none]
-let conf_venue = #group 0 conf
+let val conf_venue = #group 0 conf
 \end{lstlisting}
@@
 the positions of the captured groups.
+We would instead need to dynamically match the string against the regular expression explicitly:
+
+\begin{lstlisting}[numbers=none]
+let val conf_venue = nth 0 (match /SURL([A-Z]+) (\d\d\d\d)EURL/ conf)
+\end{lstlisting}
+Ensuring that there will not be an out-of-bounds error here would again require extrinsically reasoning about the term.
 %Refinement types are strictly a layer above the language, and thus cannot affect its dynamic semantics.
 
-A closely related problem is that type refinements do not introduce type disequalities.
-As a result, compilers cannot use the type refinements to optimize the representation of a value.
+Another consideration is that type refinements do not introduce type disequalities.
+As a result, a compiler cannot use the invariants they capture to optimize the representation of a value.
 For example, the fact that some value of type \lstinline{string} can be given the refined type \lstinline{rstring /A|B/} cannot be justification to locally alter its representation to, for example, a single bit, both because it could later be passed into a function expecting a standard string and because there are many other possible refinements.
-For regular strings, there may be limited benefit to controlling representation differentially, but in other situations, this can have significant performance ramifications.
-For example, it is quite common to justify the omission of a primitive type of ``non-negative integers representable in 32 bits'' by the fact that this is ``merely'' a refinement of integers, but in so doing one loses the ability to use a machine representation more suitable to this particular static invariant.%Refinement types cannot be used for this purpose without the possibility of conflict.
+Of course, for regular strings, there may be limited benefit (or even some cost) to controlling representation differentially, but in other situations, this could have significant performance ramifications.
+For example, it is quite common to justify the omission of a primitive type of ``non-negative integers representable in 32 bits'' by the fact that this is ``merely'' a refinement of (mathematical) integers, but in so doing the compiler loses the ability to use a machine representation more suitable to this particular static invariant.%Refinement types cannot be used for this purpose without the possibility of conflict.
 %This is also the reason why one cannot viably do away with primitive record types by claiming that they are refinements of finite mappings with labels as keys.
@@
 \begin{lstlisting}[numbers=none]
-signature TITLE
+signature TITLE = sig {
   type t
@@
   val strcase : t -> (unit -> 'a) -> (string * string -> 'a) -> 'a
-
-signature CONF
+}
+
+signature CONF = sig {
   type t
@@
   val prj_group_1 : t -> string
-\end{lstlisting}
-
-To ensure that applications of the introductory operators are correct statically, we would need to couple this with a system of refinement types.
-
-Unfortunately, this again reveals itself to be too limited a solution when we need other operators, e.g.
-concatenation.
-Expressing concatenation uniformly using a single function or a finite set of functions is impossible, for the same reason that applied to \lstinline{ltuple+} above, i.e.
-there are infinitely many regular string types.
-
-\subsection{Contributions}\label{sec:metamodules-approach}
+}
+\end{lstlisting}
+
+To ensure that applications of the introductory operators are correct statically, we would need to couple this with a system of type refinements.
+Moreover, this again reveals itself to be too limited a solution when we need to express other operators.
+For example, expressing concatenation uniformly using a single function or a finite set of functions is impossible, for the same reason as applied to \lstinline{ltuple+} above, i.e.
+there are infinitely many regular string types that might be involved.
+
+\subsection{Contributions}\label{sec:metamodules-solution}
 
@@
 
-\emph{Metamodules} introduce type and operator constructors parameterized by arbitrary \emph{static values} and metaprogrammatically express the logic governing their typechecking and  translation to the IL.
-\emph{Metasignatures} classify metamodules.
-Let us begin by considering the following metasignature, which classifies metamodules that express the semantics of labeled tuples as described above:
-\begin{figure}
+Verse introduces a \emph{metamodule system} that gives library providers more direct control over the semantics of the EL.
+Using the metamodule system, library providers can express the type structure of labeled tuples and regular strings more directly, as we will show below.
+
+Just as the Verse (i.e.
+ML) module system is organized around \emph{signatures} and \emph{modules}, the Verse metamodule system is organized around \emph{metasignatures} and \emph{metamodules}.
+Briefly summarized:
+\begin{itemize}
+\item A \emph{metasignature} specifies a set of type and operator constructors (tycons and opcons). Each is parameterized by a specified \emph{kind} of \emph{static value}. 
+\item A \emph{metamodule}  defines static functions that govern the typechecking and  translation to the IL of each type and operator constructor specified in the  metasignature. 
+\end{itemize}
+\subsubsection{Example: Labeled Tuple Types via Metamodules}
+\begin{figure}[hp!]
 \begin{lstlisting}
-metasignature LTUPLEX
-  tycon c of LblTyOrdMap
-  ana opcon intro_unlabeled (*of unit*)
-  ana opcon intro_labeled of list(Lbl)
-  syn opcon # of Lbl
-  syn opcon + (*of unit*)
-  syn opcon - of Lbl
-
-syntax ltpl(L ~ LTUPLEX, param : LblTyOrdMap) at L.c(param)
-  fn ps => (* ... intro syntax ... *)
-
-metamodule Ltuplex ~ LTUPLEX with syntax ltpl(Ltuplex) at c
-  (* translate labeled tuple types to nested products *)
-  tycon c(param : LblTyOrdMap) ~> (fold (list_of_ltom param)
+metasignature LTUPLEX = metasig {
+  ^\label{line:ltuplex-c-sig}^tycon c of LblTyOrdMap
+  ^\label{line:intro-unlabeled-sig}^ana opcon intro_unlabeled (*of unit*)
+  ^\label{line:intro-labeled-sig}^syn opcon intro_labeled of list(Lbl)
+  ^\label{line:prj-sig}^syn opcon # of Lbl
+  ^\label{line:conc-sig}^syn opcon + (*of unit*)
+  ^\label{line:drop-sig}^syn opcon - of Lbl
+}
+
+syntax ltplx(L :: LTUPLEX, param :: LblTyOrdMap) at L.c(param) {
+  fn ps => (* 
+    - elaborates {e_1, ..., e_n} to L.intro_unlabeled e_1 ... e_n
+    - elaborates {lbl1 => e_1, ..., lbln => e_n} to 
+        L.intro_labeled [lbl1, ..., lbln] e_1 ... e_n
+  *)
+}
+
+metamodule Ltuplex :: LTUPLEX = metamod {
+  ^\label{line:c-start}^(* translate labeled tuple types to nested products *)
+  tycon c {
+  	(* type translation schema: *)
+  	fn (param :: LblTyOrdMap) => fold (list_of_ltom param)
       `SCSSunitECSS`
-    (fn ((lbl, ty), cur_trans) => `SCSStrans(ECSStySCSS) * %ECSScur_trans`)
+      (fn ((lbl, ty), cur_ty_trans) => `SCSStrans(ECSStySCSS) * %ECSScur_ty_trans`)
+  }^\label{line:c-end}^
+
+  ana opcon intro_unlabeled {
+  	(* translate intro operator to nested pairs *)
+  	fn (ty :: Ty, param :: unit, args :: list(Arg)) => tycase(ty) {
+        c(param) => (fold (zipExact (list_of_ltom param) args)
+      	  `SCSS()ECSS`
+          (fn (((lbl, ty), arg), cur_trans) => `SCSS(%{ECSSana(arg, ty)SCSS}, %ECSScur_transSCSS)ECSS`)
         )
-
-  (* translate intro operator to nested pairs *)
-  ana opcon intro_unlabeled(ty : Ty, param : unit, args : list(Arg)) ~> (
-    tycase(ty) {
-      c(param) => (fold (zipExact (list_of_ltom param) args)
-      	`()`
-      	(fn (((lbl, ty), arg), cur_trans) => `(%{ana(arg)}, %cur_trans)`)
-      )
-      _ => raise TyErr("...")
-    }
-  )
-
-  (* same as intro_unlabeled but reorder first *)
-  ana opcon intro_labeled(ty : Ty, param : list(Lbl), args : list(Arg)) ~> 
-    (* ... *)
-
-  (* generate the appropriate n-fold projection out of nested pairs *)
-  syn opcon #(param : Lbl, args : list(Arg)) ~> 
-    (* ... *)
-
-  (* generate the new type ..... *)
-  syn opcon +(param : unit, args : list(Arg)) ~> 
-    (* ... *)
-  syn opcon -(param : Lbl, args : list(Arg)) ~> 
-    (* ... *)
+      | _ => raise TyErr("...")
+    }
+  }
+
+  syn opcon intro_labeled {
+  	fn (param :: list(Lbl), args :: list(Arg)) => 
+      (* ... same as intro_unlabeled but reorder first ... *)
+  }
+
+  syn opcon # {
+  	fn (param :: Lbl, args :: list(Arg)) =>  
+      (* ... generate the appropriate n-fold projection ... *)
+  }
+
+  syn opcon + {
+  	fn (param :: unit, args :: list(Arg)) => 
+      (* generate the new type and combine the two nested tuples *)
+  }
+
+  syn opcon - {fn (param : Lbl, args : list(Arg)) => 
+    (* generate the new type and drop the appropriate element *)
+  }
+} with syntax ltplx for c 
 (* synonyms used in Sec. 5.1 *)
- let tycon ltuplex = LTupleX.c
+^\label{line:tycon-synonym}^let tycon ltuplex = LTupleX.c
 let opcon # = LTupleX.#
@@
 let opcon ltuple- = LTupleX.-
-
 \end{lstlisting}
@@
 \end{figure}
-
-On line 2, we declare a tycon, identifying it as \lstinline{c}, and specify that it is parameterized by static values of type \lstinline{LblTyOrdMap}, which we assume has been defined as the type of ordered finite mappings from labels, which are values of type \lstinline{Lbl}, to types, which can be manipulated as static values of type \lstinline{Ty}.
-
-Let us assume a metamodule \lstinline{LTupleX} that implements this metasignature,
-
-c is a tycon parameterized by static maps from labels to types.
+An example of a metasignature, \lstinline{LTUPLEX}, and a matching 
+metamodule, \lstinline{Ltuplex}, is shown in Figure \ref{fig:ltuplex}.
+We also show a \emph{metamodule-parameterized TSM} \lstinline{ltplx} to decrease syntactic cost.
+The remainder of this subsection explains these definitions.
+
+\paragraph{Type Constructors} On line \ref{line:ltuplex-c-sig}, the metasignature \lstinline{LTUPLEX} specifies a type constructor \lstinline{c} parameterized by static values of kind \lstinline{LblTyOrdMap}.
+This user-defined kind (not shown here for brevity) classifies ordered mappings from labels (of kind \lstinline{Lbl}, also user-defined but not shown) to types, which are static values of kind \lstinline{Ty} (a primitive kind).
+Let us assume that we have defined a \emph{kind-specific language} for \lstinline{LblTyOrdMap}, so we can write a static value of this kind concretely like this:
+\begin{lstlisting}[numbers=none]
+let static val Paper_ty_idx :: LblTyOrdMap = {
+  title : rstring /.+/
+   conf : rstring /([A-Z]+) (\d\d\d\d)/
+}
+\end{lstlisting}
+
+We can construct a type by applying the type constructor \lstinline{Ltuplex.c} to this parameter:
+\begin{lstlisting}[numbers=none]
+let type Paper = Ltuplex.c Paper_ty_idx
+\end{lstlisting}
+Line \ref{line:tycon-synonym} defines the tycon synonym \lstinline{ltuplex} for \lstinline{Ltuplex.c}.
+If we also substitute in the value of \lstinline{Paper_ty_idx} just given, we see that we have recovered the definition of \lstinline{Paper} given at the beginning of Sec.
+\ref{sec:metamodules-example}:
+\begin{lstlisting}[numbers=none]
+let type Paper = ltuplex {
+  title : rstring /.+/
+   conf : rstring /([A-Z]+) (\d\d\d\d)/
+}
+\end{lstlisting}
+Note that because types are static values, this is equivalent to writing:
+\begin{lstlisting}[numbers=none]
+let static val Paper :: Ty = ltuplex {
+  title : rstring /.+/
+   conf : rstring /([A-Z]+) (\d\d\d\d)/
+}
+\end{lstlisting}
+
+Recall from Sec.
+\ref{sec:verse} that each valid external type has a translation, which is an internal type.
+The metamodule \lstinline{Ltuplex} defines a \emph{type translation schema}, a static function that determines the translations of types constructed by \lstinline{c} (each identified by the value of its parameter).
+Internal types are encoded as static values of type \lstinline{ITy}, which we assume has a TSL associated with it that allows us to use standard \emph{quasiquotation} syntax.
+Here, the  type translation schema on lines \ref{line:c-start} to \ref{line:c-end} generates nested internal products by folding over a list generated from the type parameter (in the standard way), returning the encoding of \lstinline{unit} for the empty case, and recursively nesting encodings of binary products otherwise using the standard \emph{unquote} operator, \lstinline{%}.
+For example, the encoding of the type translation of \lstinline{Paper} determined by this static function is:
+\begin{lstlisting}[numbers=none]
+`SCSStrans(ECSSrstring /SURL.+EURL/SCSS) * (trans(ECSSrstring /SURL([A-Z]+) (\d\d\d\d)EURL/SCSS) * unit)ECSS`
+\end{lstlisting}
+Notice that the translations of the component regular string types are not inserted directly, but are rather treated \emph{parametrically} using the \emph{translational form} \lstinline{SCSStrans(ECSSTSCSS)}.
+This will be key to the modularity principle that we plan to explore in the remaining work.
+
+If we assume that all regular string types translate to standard internal strings, then the translation of \lstinline{Paper} is ultimately:
+\begin{lstlisting}[numbers=none]
+string * (string * unit)
+\end{lstlisting}
+
+\paragraph{Operator Constructors} On line \ref{line:intro-unlabeled-sig}, the metasignature \lstinline{LTUPLEX} specifies an operator constructor \lstinline{intro_unlabeled}.
+No parameter kind is listed, so the trivial parameter kind, \lstinline{unit}, is assumed, as indicated by the comment.
 
@@
 \end{lstlisting}
+
+\subsection{Timeline}\label{sec:metamodules-timeline}
 % The semantics uses internal type abstraction to guarantee that metamodule providers can reason modularly about translation invariants, just as it uses external type abstraction to guarantee that module providers can reason modularly about representation invariants.%For example, a metamodule that expresses the type structure of records wouldLblTyList introduce a type constructor  \texttt{record} parameterized by a static mapping from labels to types, and two operators: the introductory operator \texttt{intro}, parameterized by a list of row labels, and the row projection operator \texttt{prj}, parameterized by a row label.

### 2015-04-28 thesis-proposal/problem.tex
update

@@
   ^\label{line:ltuplex-c-sig}^tycon c of LblTyOrdMap
-  ^\label{line:intro-unlabeled-sig}^ana opcon intro_unlabeled (*of unit*)
+  ^\label{line:intro-unlabeled-sig}^ana opcon intro_unlabeled of unit
   ^\label{line:intro-labeled-sig}^syn opcon intro_labeled of list(Lbl)
   ^\label{line:prj-sig}^syn opcon # of Lbl
-  ^\label{line:conc-sig}^syn opcon + (*of unit*)
+  ^\label{line:conc-sig}^syn opcon + of unit
   ^\label{line:drop-sig}^syn opcon - of Lbl
@@
   fn ps => (* 
-    - elaborates {e_1, ..., e_n} to L.intro_unlabeled e_1 ... e_n
+    - elaborates {e_1, ..., e_n} to L.intro_unlabeled(e_1; ...; e_n)
     - elaborates {lbl1 => e_1, ..., lbln => e_n} to 
@@
 
-  ana opcon intro_unlabeled {
+  ^\label{line:introul-start}^ana opcon intro_unlabeled {
   	(* translate intro operator to nested pairs *)
@@
     }
-  }
+  }^\label{line:introul-end}^
 
@@
 \end{lstlisting}
-Notice that the translations of the component regular string types are not inserted directly, but are rather treated \emph{parametrically} using the \emph{translational form} \lstinline{SCSStrans(ECSSTSCSS)}.
+Notice that the translations of the component regular string types are not inserted directly, but are rather treated {parametrically} using the \emph{translational form} \lstinline{SCSStrans(ECSSTSCSS)}.
 This will be key to the modularity principle that we plan to explore in the remaining work.
@@
 
-\paragraph{Operator Constructors} On line \ref{line:intro-unlabeled-sig}, the metasignature \lstinline{LTUPLEX} specifies an operator constructor \lstinline{intro_unlabeled}.
-No parameter kind is listed, so the trivial parameter kind, \lstinline{unit}, is assumed, as indicated by the comment.
-
-intro unlabeled is a trivially parameterized opcon, i.e. just one operator.
-
-intro labeled is an opcon parameterized by a list of labels.
-
-\# is an opcon parameterized by a single label.
-
-ltpl is a TSM parameterized by a metamodule and a static value
-
-LTuple uses ltpl as a TSL
-
-\begin{lstlisting}
-\end{lstlisting}
+\paragraph{Operator Constructors} On line \ref{line:intro-unlabeled-sig} of Figure \ref{fig:ltuplex}, the metasignature \lstinline{LTUPLEX} specifies an operator constructor, \lstinline{intro_unlabeled} parameterized by static values of kind \lstinline{unit}.
+The qualifier \lstinline{ana} specifies that this opcon can only be used in analytic positions (i.e.
+where the expected type is known).
+For example, \lstinline{Ltuplex.intro_unlabeled} can be applied like this, because the return type of the function determines the expected type:
+\begin{lstlisting}[numbers=none]
+fun make_paper(conf : conf_t, title : title_t) : Paper => 
+  Ltuplex.intro_unlabeled () (conf; title)
+\end{lstlisting}
+First, a static parameter value of the appropriate kind is supplied. After that, an \emph{argument list}, \lstinline{(conf; title)}, is supplied.
+
+The metamodule \lstinline{Ltuplex} determines how this expression is typechecked and translated on lines \ref{line:introul-start}-\ref{line:introul-end}.
+Analytic opcons like \lstinline{intro_unlabeled} are defined with a static function that is given the type that the expression is being analyzed against, here \lstinline{Paper}, the operator parameter, here \lstinline{()}, and a list of \emph{argument interfaces}, which will allow it to ask the semantics to recursively typecheck and translate the provided arguments.
+The function must return an encoding of an  internal expression, which will become the translation of the external expression.
+Internal expressions are encoded as static values of kind \lstinline{IExp}, and again we assume a TSL has been defined so that we can use standard quasiquote syntax.
 

### 2015-05-06 thesis-proposal/problem.tex
almost done...

@@
 Functional programming languages like Standard ML (SML), OCaml and Haskell descend from mathematically elegant typed lambda calculi, diverging mainly in that they consider various pragmatic issues alongside  fundamental metatheoretic issues in their design.
-For example, they build in record types, which generalize the nullary and binary product types more common in minimal calculi, because labeled components can be cognitively useful to humans.
-Similarly, they build in derived syntax for commonly used constructs like lists.% concision and readability.
+For example, they build in record types, generalizing the nullary and binary product types more common in minimal calculi, because labeled components are often cognitively useful.
+Similarly, they build in derived concrete syntax for commonly used constructs like lists.% concision and readability.
 %Minimal typed lambda calculi play a central role in the study of programming languages, and serve as direct inspiration for  functional programming languages like ML.
@@
 
-Ideally, a limited number of ``embellishments'' like these would suffice to transition from a minimal calculus to a practical ``full-scale'' programming language.
-The community around the resulting language would then be able to direct its efforts exclusively toward the development of useful libraries.
-Unfortunately, a stable language design that achieves this ideal has yet to emerge, as evidenced by the diverse array of {dialects} that continue to proliferate around all major contemporary languages.
+Ideally, only a limited number of ``embellishments'' like these would be needed to turn a minimal lambda calculus into a practical ``full-scale'' functional programming language.
+The community around the resulting language would then be able to direct its efforts almost exclusively toward the development of useful libraries, optimizing compilers and other tools.
+Unfortunately, a stable language design that fully achieves this ideal has yet to emerge, as evidenced by the diverse array of \emph{dialects} -- languages derived from a common base language that introduce different embelishments -- that continue to proliferate around all major contemporary languages.
 % For example, let us consider the many dialects of ML.
@@
 %Tools for  constructing new ``domain-specific'' dialects also continue to proliferate. % The situation is similar across all major language lineages. 
-Indeed, tools that assist with the construction of  ``domain-specific'' language dialects seem only to be increasingly popular. 
-{This calls for an investigation}: why are programmers and researchers still so often unable to satisfyingly express the constructs that they need in libraries, as modes of use of the ``general-purpose'' primitives already available in major functional languages?
+In fact, tools that assist with the construction of so-called  ``domain-specific'' language dialects seem only to be increasingly popular. 
+{This calls for an investigation}: why are programmers and researchers so often unable to satisfyingly express the constructs that they need in libraries, as modes of use of the ``general-purpose'' primitives already available in major functional languages?
 %Let us investigate why this is.
@@
 %Syntactic cost is often assessed qualitatively \cite{green1996usability}, though quantitative metrics can be defined.
-Many other types of data can similarly benefit from the availability of specialized derived syntax.
+Many other types of data similarly benefits from the availability of specialized derived syntax.
 For example, we will consider regular expression patterns expressed using abstract data types (as a mode of use of the ML module system) in Sec.
 \ref{sec:syntax}.
-%Though we will be able to express  precisely the semantics that we seek,   the usual approach of using dynamic string parsing to introduce a pattern will leave something to be desired syntactically.
+Though we will be able to express  precisely the semantics that we seek,   the usual approach of using dynamic string parsing to introduce a pattern will leave something to be desired syntactically.
 % An extension of Moscow ML provided syntax for monads encoded using the module system, inspired by the  syntax built in to Haskell.\todo{asked for reference, if not found will say that "Harper reports..."} %In fact, nearly all major programming languages primitively build in derived syntax for constructs that can otherwise be expressed in the standard library, e.g.
@@
 Record types themselves are perhaps  the simplest example -- they cannot be expressed by context-free elaboration to a language with only nullary and binary products.
-Various dialects of ML and Haskell have explored  ``record-like'' primitives that go even further, supporting functional update and extension operators, width and depth coercions (sometimes implicit)%\cite{Cardelli:1984:SMI:1096.1098}
-, methods, prototypic dispatch and other semantic ``embellishments'' that cannot be expressed by context-free elaboration to a language with only standard record types (we will detail one  example in Sec.
+Various dialects of ML and Haskell have explored  ``record-like'' primitives that go still further, supporting functional update and extension operators, width and depth coercions (sometimes implicit)%\cite{Cardelli:1984:SMI:1096.1098}
+, methods, prototypic dispatch and other ``semantic embellishments'' that cannot be expressed by context-free elaboration to a language with only standard record types (we will detail one  example in Sec.
 \ref{sec:metamodules}).
@@
 As MLj demonstrates, addressing the interoperability problem requires somehow ``combining'' the  dialects into a single language.
-In the most general setting where the dialects in question can be specified by judgements of arbitrary form, this is not a well-defined notion.
+In the most general setting where the dialects in question might be specified by judgements of arbitrary form, this is not a well-defined notion.
 Even if we restrict our interest to dialects specified using formalisms that do operationalize some notion of dialect combination, there is generally no guarantee that the combined dialect will conserve syntactic and semantic properties that can be established about the dialects in isolation.
@@
 %Over time, accepting such extensions has caused these languages to balloon in size. 
-This \emph{ad hoc} approach is not sustainable, for two main reasons.
+This \emph{ad hoc} approach is not sustainable, for three main reasons.
 First, as suggested by the diversity of examples given above, there are simply too  many potentially useful such primitives, and many of these are only relevant in relatively narrow application domains (for derived syntax, our group has  gathered initial data speaking to this \cite{TSLs}).
 Second, primitives introduced earlier in a language's lifespan end up monopolizing finite ``syntactic resources'', forcing subsequent primitives to use ever more esoteric forms.
-A related problem is that primitives cannot be removed or changed without breaking backwards compatibility.
+A third problem is that primitives cannot be removed or changed without breaking backwards compatibility.
 %rimitives that are only situationally useful, or that trigger aesthetic disagreements between different factions of the community, are still quite often left languishing in impractical ``toy'' dialects.
@@
 and generality in language design is a serious unsolved 
-problem.}%\begin{flushright}--- John Reynolds (1970)\end{flushright}
+problem.}\begin{flushright}--- John Reynolds (1970)\end{flushright}
 \end{quote}
 
-\newpage %A measured approach to incorporating new primitives into a general-purpose programming language is  sensible because once a primitive is introduced, it becomes entrenched as-is and monopolizes ``syntactic resources'', as we will discuss below.
+%A measured approach to incorporating new primitives into a general-purpose programming language is  sensible because once a primitive is introduced, it becomes entrenched as-is and monopolizes ``syntactic resources'', as we will discuss below.
 This leaves the subset of the  language design community interested in keeping general-purpose languages small and free of \emph{ad hoc} primitives with two possible paths forward.
@@
 % that this implies.
-The other is to search for even more general primitives that reduce these seemingly \emph{ad hoc} primitives to modularly composable library constructs.
+The other is to search for more general primitives that reduce these seemingly \emph{ad hoc} primitives to modularly composable library constructs.
 % so that they can be evaluated on their individual merits by programmers and used together without the possibility of conflict.
@@
 For example, a recent revision of Ocaml added support for  ``generalized algebraic data types'' (GADTs), based on research on guarded recursive datatype constructors \cite{XiCheChe03}.
-Using GADTs, Ocaml was able to move some of the \emph{ad hoc} machinery for typechecking operations that use format strings, like \texttt{sprintf}, out of the language and into a library (however, syntactic machinery remains built in).
+Using GADTs, Ocaml was able to move some of the \emph{ad hoc} machinery for typechecking operations that use format strings, like \texttt{sprintf}, out of the language and into a library (note however that syntactic machinery remains built in).
 
@@
 \section{Proposed Contributions}
-Our broad aim in the work being proposed is to introduce a new functional programming language called Verse\footnote{We distinguish Verse from Wyvern, which is the language referred to in prior publications about some of the work being proposed here, because Wyvern is a group effort evolving independently in some important ways.} that takes more radical steps down the second path just described by giving library providers the ability to directly express derived syntax and type structure of a variety of designs in a safe and modularly composable manner.
+Our broad aim in the work being proposed is to introduce a new functional programming language called Verse\footnote{We distinguish Verse from Wyvern, which is the language referred to in prior publications about some of the work being proposed here, because Wyvern is a group effort evolving independently in some important ways.} that takes more radical steps down the second path just described by giving library providers the ability to directly express derived concrete syntax and type structure of a variety of designs in a safe and modularly composable manner.
 %This eliminates the needs to build in fewer \emph{ad hoc} constructs and dialects are less frequently necessary.
@@
 \begin{enumerate}
-\item Introduces a type constructor, \lstinline{record}, parameterized by finite mappings from labels to types and defines its semantics by translation to unary and binary products (which are built in to the internal language).
-\item Introduces operators used to work with records, minimally record introduction and elimination (but perhaps also various functional update operators), and directly defines the logic governing their typechecking and translation.
+\item introduces a type constructor, \lstinline{record}, parameterized by finite mappings from labels to types, and defines its semantics by translation to unary and binary products (which are built in to the internal language); and
+\item introduces operators used to work with records, minimally record introduction and elimination (but perhaps also various functional update operators), and directly defines the logic governing their typechecking and translation to the IL (which builds in only nullary and binary products).
 \end{enumerate}
@@
 \end{itemize} 
-Both TSMs and metamodules can be understood as \emph{metaprogramming} primitives, because their static semantics will involve the evaluation of user-defined functions that manipulate static representations of types and terms.% static code generation.
+Both TSMs and metamodules involve \emph{static metaprogramming}, i.e.
+their static semantics will involve the evaluation of user-defined functions that manipulate static representations of types and terms.% static code generation.
 %We will  also introduce a simple variant of each of these primitives that leverages Verse's support for local type inference to further reduce syntactic cost in certain common situations.
 
-The key challenge in the design of these primitives will come in ensuring that they are metatheoretically well-behaved, given that they aim to decentralize control over aspects of the language's concrete syntax and type structure that, in other contemporary languages, are under the exclusive control of the language designer.
-If we are not careful, many of the problems discussed earlier that arise when combining language dialects could simply shift into the semantics of these primitives.\footnote{This is why languages  like Verse are often called \emph{extensible languages}, though this is somewhat of a misnomer.
+The key challenge in the design of these primitives will come in ensuring that they are metatheoretically well-behaved, given that they aim to decentralize control over aspects of a language's concrete syntax and type structure that, in other contemporary languages, are under the exclusive control of the language designer.
+If we are not careful, many of the problems  that arise when combining language dialects, discussed earlier, could simply shift into the semantics of these primitives.\footnote{This is why languages  like Verse are often called \emph{extensible languages}, though this is somewhat of a misnomer.
 The chief characteristic of an extensible language is that it \emph{doesn't} need to be extended in situations where other languages would need to be extended.
@@
 In particular, syntactic conflicts will be impossible by construction and the semantics will validate code statically generated by TSMs and metamodules to maintain a strong \emph{hygienic type discipline} and, most uniquely, powerful \emph{modular reasoning principles}.
-In other words, library providers will have the ability to reason about the constructs that they have defined in isolation, and clients will be able to use them safely in any program context and in any combination, without the possibility of conflict.\footnote{This is not quite true -- simple naming conflicts can arise.
+In other words, library providers will have the ability to reason about the constructs that they have defined in isolation, and clients will be able to use them safely in any program context (i.e.
+\emph{hygenically}) and in any combination, without the possibility of conflict (i.e.
+\emph{modularly}).\footnote{This is not quite true -- simple naming conflicts can arise.
 We will tacitly assume that they are being avoided extrinsically, e.g.
@@
 \subsection{Disclaimers}
-Before we continue, it may be useful to explicitly acknowledge that completely eliminating the need for dialects would indeed be asking for too much: certain design decisions are fundamentally incompatible with others or require coordination across a language design.
+Before we continue, let us explicitly acknowledge that completely eliminating the need for dialects would indeed be asking for too much: certain design decisions are fundamentally incompatible with others or require coordination across a language design.
 We aim only to decrease the need for dialects.% out a larger design space within a single language, Verse.%a subset of constructs that can be specified by a semantics of a certain ``shape'' specified by Verse (we will make this more specific later).
@@
 
-Finally, we will not be interested here in languages that feature full-spectrum dependent types, which blur the phase separation between compile-time and run-time, though we conjecture that the primitives we introduce could be introduced into languages like Gallina (the ``external language'' of the Coq proof assistant) with some reasonable modifications.
+Finally, we will not be interested here in languages that feature full-spectrum dependent types, which blur the phase separation between ``compile-time'' and ``run-time'', though we conjecture that the primitives we describe could be added to languages like Gallina (the ``external language'' of the Coq proof assistant) with some suitable modifications.
 Verse is designed for use in situations where ML, Haskell or Scala would be suitable today.
@@
 
-The Verse module language is based directly on SML's module language, with which we assume a working familiarity for the purposes of this proposal \cite{harper1997programming,MacQueen:1984:MSM:800055.802036} (a related module language, e.g.
+\subsection{Module Language}
+The Verse module language is taken directly from Standard ML, with which we assume a working familiarity for the purposes of this proposal \cite{harper1997programming,MacQueen:1984:MSM:800055.802036} (a related module language, e.g.
 Ocaml's, would work just as well for our purposes).
@@
 
-The Verse core language will be the focus of our formal efforts.
-It is organized much like the first stage of a type-directed compiler (e.g.
+\subsection{Core Language}
+The underlying core language -- the language of types and expressions -- will be the focus of our formal efforts.
+The Verse core language is organized much like the first stage of a type-directed compiler (e.g.
 the TIL compiler for Standard ML \cite{tarditi+:til-OLD}), consisting of a user-facing \emph{typed external language} (EL) specified by type-directed translation to a minimal \emph{typed internal language} (IL).
-The main judgements in the specification of the EL take the following form  (omitting various contexts for now):\footnote{Note that we will use the terms ``expression'' and ``type'' without qualification to refer to  external expressions and types, respectively.
-}
-\\[1ex]
-$\begin{array}{ll}
-\textbf{Judgement Form} & \textbf{Pronunciation}\\
-\vdash e \Rightarrow \sigma \leadsto \iota & \text{Expression $e$ synthesizes type $\sigma$ and has translation $\iota$.}\\
-\vdash e \Leftarrow \sigma \leadsto \iota & \text{Expression $e$ analyzes against type $\sigma$ and has translation $\iota$.}\\
-\vdash \sigma~\mathtt{type} \leadsto \tau & \text{Static expression $\sigma$ is a type with translation $\tau$.}
-\end{array}
-$\\[1ex]
-Note that the expression typing judgements are \emph{bidirectional}, i.e.
-we make a judgemental distinction between \emph{type synthesis} (the type is an ``output'') and \emph{type analysis} (the type is an ``input'') \cite{Pierce:2000:LTI:345099.345100}.
-This is to  allow us to explicitly specify how Verse's \emph{local type inference} works, and in particular, how it interacts with the primitives that we aim to introduce.
-Like Scala, we intentionally do not attempt to support global type inference.%, primarily because it is not technically compatible with the primitives we will introduce (but also because we believe that compilers for languages that support only local type inference generate clearer error messages).
-
-% The reason why we refer to the syntactic sort $\sigma$ as a ``static expression'' will become more clear later.
-
-%The main novelty of our work is that the Verse EL builds in very little syntax or type structure \emph{a priori}.
-%Instead, \textbf{metamodules} and \textbf{typed syntax macros} (TSMs) are expressive enough to render this unnecessary.
-Although metamodules are in some sense prior to TSMs, we will describe TSMs first, temporarily assuming that the EL does build in standard type structure, for pedagogical purposes.
-
-Our design is largely insensitive to the details of the internal language -- the only strict requirements are that the IL be type safe and support parametric type abstraction.
-For our purposes, we will keep things simple by using the strictly evaluated polymorphic lambda calculus with nullary and binary product and sum types and recursive types, which we assume the reader has familiarity with and for which all the necessary metatheorems are well-established (we follow \cite{pfpl}).
-The main judgements in the static semantics of the IL (again omitting contexts for now) take the following standard form:
+
+\subsubsection{Internal Language}
+Let us begin with the internal language.
+For our purposes, we keep things simple by using the strictly evaluated polymorphic lambda calculus with nullary and binary product and sum types and recursive types.
+We assume here that the reader has familiarity with these (we follow \emph{PFPL} \cite{pfpl} directly).
+To briefly review, the main judgements in the static semantics of the IL take the following standard form (omitting contexts for now):
 \\[1ex]
@@
 
-Features like state, exceptions, simple concurrency primitives, scalars, arrays and others characteristic of a first-stage compiler intermediate language would also be included in the IL in practice, and in some cases this would affect the shape of the internal semantics, but we will omit them for simplicity.
+Features like state, exceptions, minimal concurrency primitives, scalars, arrays and others characteristic of a first-stage compiler intermediate language would also be included in the IL in practice, and in some of these cases, this would affect the shape of the internal semantics.
+However, our design is largely insensitive to the details of the internal language -- the only strict requirements are that the IL be type safe and support parametric type abstraction -- so we will stick to this simpler core here.
 %Features like record types and labeled sum types, on the other hand, are not built in.
 
+\subsubsection{External Language}
+Programs evaluate as internal expressions but they are written as external expressions.
+For concision, we will use the terms ``type'' and ``expression'' without qualification to refer to  \emph{external types}, $\sigma$,  and \emph{external expressions}, $e$, respectively.
+Note that external types are distinct from internal types.
+The main judgements in the specification of the EL take the following form  (again omitting various contexts for now):
+\\[1ex]
+$\begin{array}{ll}
+\textbf{Judgement Form} & \textbf{Pronunciation}\\
+\vdash e \Rightarrow \sigma \leadsto \iota & \text{Expression $e$ synthesizes type $\sigma$ and has translation $\iota$.}\\
+\vdash e \Leftarrow \sigma \leadsto \iota & \text{Expression $e$ analyzes against type $\sigma$ and has translation $\iota$.}\\
+\vdash \sigma~\mathtt{type} \leadsto \tau & \text{Static expression $\sigma$ is a type with translation $\tau$.}
+\end{array}
+$\\[1ex]
+Note that the expression typing judgements are \emph{bidirectional}, i.e.
+we make a judgemental distinction between \emph{type synthesis} (the type is an ``output'') and \emph{type analysis} (the type is an ``input'') \cite{Pierce:2000:LTI:345099.345100}.
+This is to  allow us to explicitly specify how Verse's \emph{local type inference} works, and in particular, how it interacts with the primitives that we aim to introduce.
+Like Scala, we intentionally do not attempt to support global type inference.
+%, primarily because it is not technically compatible with the primitives we will introduce (but also because we believe that compilers for languages that support only local type inference generate clearer error messages).
+
+% The reason why we refer to the syntactic sort $\sigma$ as a ``static expression'' will become more clear later.
+
+%The main novelty of our work is that the Verse EL builds in very little syntax or type structure \emph{a priori}.
+%Instead, \textbf{metamodules} and \textbf{typed syntax macros} (TSMs) are expressive enough to render this unnecessary.
+Although metamodules are in some sense prior to TSMs, we will describe TSMs first, temporarily assuming that the EL does build in standard type structure, for pedagogical purposes.
+
 \section{Modularly Metaprogrammable Textual Syntax}\label{sec:syntax}
-Verse, like most major contemporary programming languages, specifies a textual concrete syntax.\footnote{Although Wyvern specified a layout-sensitive concrete syntax, to avoid unnecessary distractions, we will specify a more conventional layout-insensitive concrete syntax for Verse.} %We have chosen to specify a layout-sensitive textual concrete syntax (i.e.
+Verse specifies a textual concrete syntax for the EL.\footnote{Although Wyvern specified a layout-sensitive concrete syntax, to avoid unnecessary distractions, we will specify a more conventional layout-insensitive concrete syntax for Verse.} %We have chosen to specify a layout-sensitive textual concrete syntax (i.e.
 newlines and indentation are not ignored).
@@
 %For constructs that have an obvious analog in ML, we will omit a detailed explanation.
-Because the purpose of concrete syntax is to serve as the language's programmer-facing user interface, it is common practice to build in  derived syntactic forms (colloquially, \emph{syntactic sugar}) that capture common idioms more concisely or naturally.
+Because the purpose of concrete syntax is to serve as a programmer-facing user interface, it is common practice to build in  derived syntactic forms (colloquially, \emph{syntactic sugar}) that capture common idioms more concisely or naturally.
 % (i.e.
 considering cognitive dimensions \cite{green1996usability}).
-For example, derived list syntax is built in to most functional languages, so that instead of having to write \lstinline{Cons(1, Cons(2, Cons(3, Nil)))}, the programmer can equivalently write \lstinline{[1, 2, 3]}.
-Many languages go further, building in derived syntax associated with various other types of data, like vectors (SML/NJ), arrays (Ocaml), monadic commands (Haskell), syntax trees (Scala), XML documents (Scala, Ur/Web) and SQL queries (F\#, Ur/Web).
+For example, derived list syntax is built in to most functional languages, so that instead of having to write out \lstinline{Cons(1, Cons(2, Cons(3, Nil)))}, the programmer can equivalently write \lstinline{[1, 2, 3]}.
+Many languages go beyond this, building in derived syntax associated with various other types of data, like vectors (the SML/NJ dialect of SML), arrays (Ocaml), monadic commands (Haskell), syntax trees (Scala), XML documents (Scala, Ur/Web) and SQL queries (F\#, Ur/Web).
 %This is a rather \emph{ad hoc} process.% discussed previously, the usual approach is to require that the language designer build in new derived syntactic forms.
@@
 
-Verse takes a less \emph{ad hoc} approach -- rather than privileging  particular library constructs with primitive syntactic support, Verse exposes primitives that allow library providers to introduce new derived syntax of a variety of designs on their own.
+Verse takes a less \emph{ad hoc} and more modular approach -- rather than privileging  particular library constructs with primitive syntactic support, Verse exposes primitives that allow library providers to introduce new derived syntax of a variety of designs on their own.
 %Lists need no special consideration from the language specification.
@@
 %For forms with a clear analogy to a form in Standard ML, we will assume the  semantics are analagous without providing details.
+
 We will begin in Sec.
-\ref{sec:examples} by considering a representative example in detail: regular expression patterns expressed using abstract data types.
-We show how the usual approach of using strings to introduce patterns is less than ideal.
-We then survey existing alternatives in Sec.
-\ref{sec:syntax-existing}, finding that they involve an unacceptable loss of modularity and other undesirable trade-offs.
+\ref{sec:examples} by considering in detail a representative example for which such a mechanism might be useful: regular expression patterns expressed using abstract data types.
+In Sec.
+\ref{sec:syntax-existing}, we will demonstrate that the usual approach of using dynamic string parsing to introduce patterns is not ideal, and then survey existing alternatives, finding that they involve an unacceptable loss of modularity and other undesirable trade-offs.
 In Sec.
@@
 Recall that regular expressions are a common way to capture patterns in strings \cite{Thompson:1968:PTR:363347.363387}.
-We will assume that the abstract syntax of {patterns}, $p$, over strings, $s$, is specified as below:\[p ::= \textbf{empty} ~|~ \textbf{str}(s) ~|~ \textbf{seq}(p; p) ~|~ \textbf{or}(p; p) ~|~ \textbf{star}(p) ~|~ \textbf{group}(p)\]
+The abstract syntax of {patterns}, $p$, over strings, $s$, is specified as below:\[p ::= \textbf{empty} ~|~ \textbf{str}(s) ~|~ \textbf{seq}(p; p) ~|~ \textbf{or}(p; p) ~|~ \textbf{star}(p) ~|~ \textbf{group}(p)\]
 One way to express this abstract syntax is by defining a recursive sum type \cite{pfpl}.
@@
 Indeed, there may be many ways to represent regular expression patterns, each with different performance trade-offs.
-For these reasons, a better approach in Verse, as in ML, is to abstract over the choice of representation using  the module system.
-In particular, we can define the following \emph{module signature}, where the type of patterns, \lstinline{t}, is held abstract and functions are provided in lieu of datatype constructors and case analysis:
+For these reasons, a better approach in Verse, as in ML, is to abstract over the choice of representation using  the module system's support for type abstraction.
+In particular, we can define the following \emph{module signature}, where the type of patterns, \lstinline{t}, is held abstract:
 %Notice that it exposes an interface otherwise  to the one available using a case type:
@@
 \end{lstlisting}
- A client of a module \lstinline{P} that matches \lstinline{PATTERN}, which we write \lstinline{P : PATTERN}, can express patterns as terms of type \verb|P.t|.
-By holding the representation type of patterns abstract, the burden of proving that the case analysis function cannot be used to distinguish patterns with the same normal form is localized to each module implementing this signature.
-The details are standard and not particularly relevant to our purposes, so we omit them here.
+ Any client of a module \lstinline{P} that satisfies \lstinline{PATTERN}, written \lstinline{P : PATTERN}, can manipulate patterns as terms of type \verb|P.t| using the interface described by this signature.
+By holding the representation type of patterns abstract, the burden of proving that the case analysis function cannot be used by clients to distinguish patterns with the same normal form is localized to each module implementing this signature.
+The details are standard and not particularly relevant for our purposes, so we omit them here.
 
@@
 
-To express the concrete syntax of patterns, regular expression library providers usually provide a utility function that transforms strings to patterns.
-Because, as just mentioned, there  may be many implementations of the \lstinline{PATTERN} signature, the standard approach is to define a parameterized module (a.k.a.
-a \emph{functor} in SML) defining utility functions like this abstractly:
+\subsection{Existing Approaches}\label{sec:syntax-existing}
+\subsubsection{Dynamic String Parsing}
+To expose this more concise concrete syntax for regular expression patterns to clients, the most common approach is to provide a parse function that transforms strings to  patterns.
+Because, as just mentioned, there  may be many implementations of the \lstinline{PATTERN} interface, the standard approach is to define a parameterized module (a.k.a.
+\emph{functor} in SML) defining utility functions like this abstractly:
 
@@
 When compiling an expression like this, the programmer would see an error message like \verb|error: illegal escape character|.\footnote{This is the error message that \texttt{javac} produces.
-When compiling an analagous expression using SML of New Jersey (SML/NJ), we encounter the rather bizarre error message \texttt{Error: unclosed string}.} In a small lab study, we observed that this sort of error was common, and that diagnosing the problem was nearly always initially challenging to even experienced programmers who had not used regular expressions recently \cite{Omar:2012:ACC:2337223.2337324}.
-One inelegant workaround is to  double backslashes:
+When compiling an analagous expression using SML of New Jersey (SML/NJ), we encounter the rather bizarre error message \texttt{Error: unclosed string}.} In a small lab study, we observed that this sort of error was common, and that diagnosing the problem nearly always required a non-trivial amount of time for even experienced programmers who had not used regular expressions recently \cite{Omar:2012:ACC:2337223.2337324}.
+One workaround with higher syntactic cost is to  double backslashes:
 \begin{lstlisting}[numbers=none]
@@
 We might instead attempt to treat the well-formedness of patterns constructed from strings as a static verification condition.
-Statically proving that this condition holds throughout a program is wildly difficult to automate, because it involves reasoning equationally about arbitrary dynamic behavior.
-In the example above, we must know that the variable \lstinline{pattern} is equal to the function \lstinline{PU.parse}.
+Automatically proving that this condition holds throughout a program is difficult to do in general, because it involves reasoning about arbitrary dynamic behavior.
+Here, the decision procedure must know that the variable \lstinline{pattern} is equal to the function \lstinline{PU.parse}.
 If the argument had not been written literally but rather computed, e.g.
-if we had written \lstinline{"SSTR(GESTR" ^ "SSTRCESTR"} where \lstinline{^} is the string concatenation function applied in infix style, we would need to be able to establish that this expression is equivalent to the string above.
-For patterns that are dynamically constructed based on input to the program (further discussed below), the problem is harder still.
-
-\item Parsing patterns dynamically also incurs a performance cost. If the programmer is not careful, they might  incur this cost many times. For example:
-\begin{lstlisting}[numbers=none]
-map exmpl_list (fn x => rx_match (pattern "SSTRA|T|G|CESTR") x)
-\end{lstlisting}
-To avoid incurring the parsing cost for each element of \lstinline{exmpl_list}, the programmer or compiler must move the parsing step out of the transformation or ensure that an appropriately tuned caching strategy is being used.
-%The author's experience in bioinformatics suggests that often, none of these strategies is used.
+if we had written \lstinline{"SSTR(GESTR" ^ "SSTRCESTR"} where \lstinline{^} is the string concatenation function applied in infix style, we would need to be able to establish that this expression is equivalent to the string \lstinline{"SSTR(GCESTR"}.
+For patterns that are dynamically constructed based on input to a function (further discussed below), the problem is harder still.
+Asking the client to provide a proof of well-formedness would defeat the purpose of providing a concise concrete syntax.
+
+\item Parsing patterns dynamically also incurs a performance cost.
+Regular expression patterns are often used on large datasets, so it is easy to incur this cost repeatedly.
+For example, consider mapping over a list of strings:
+\begin{lstlisting}[numbers=none]
+map exmpl_list (fn s => rx_match (pattern "SSTRA|T|G|CESTR") s)
+\end{lstlisting}
+To avoid incurring the parsing cost for each element of \lstinline{exmpl_list}, the programmer or compiler must move the parsing step out of the transformation or ensure that an appropriately tuned caching strategy is being used.\footnote{Anecdotally, based on my experience in bioinformatics, the most common situation is that the compiler is not generally willing to partially evaluate expensive computations like pattern parsing out of such expressions automatically, and the programmer is often unaware of the performance cost.} %If the programmer does it, it can sometimes make the program more difficult to read.
+
 %This too is difficult if a portion of the pattern is dynamically generated.
@@
 
-\item The final problem is that using strings to construct patterns encourages programmers to use  operations over strings, like concatenation, to construct patterns derived from data that happens also to be of type \lstinline{string}.
-This can lead to subtle but serious bugs.
-For example, consider the following function:
+\item The next problem is that dynamic string parsing only decreases the cost of statically known patterns, but not those constructed using other values.
+For example, consider this function from strings to patterns:
+\begin{lstlisting}[numbers=none]
+fun example(name : string) => 
+  P.Seq(P.Str(name), P.Seq(pattern "SSTR: ESTR", ssn)) (* ssn as above *)
+\end{lstlisting}
+We needed to use both dynamic string parsing and explicit applications of pattern constructors to achieve the intended semantics.
+It is difficult to capture idioms like this more concisely using dynamic string parsing (we will see more concise syntax below).
+
+\item Finally, for functions like the one just described where we are constructing patterns on the basis of data of type \lstinline{string}, it is easy for programmers to make  mistakes  using string concatenation.
+For example, consider the following ``alternative'' definition of \lstinline{example}:
 \begin{lstlisting}[numbers=none,escapechar=|]
@@
 \end{lstlisting}
-The (unstated) intent here is to treat \lstinline{name} as a sub-pattern matching only itself, but this is not the observed behavior when \lstinline{name} contains special characters that have other meanings in patterns.
-The correct code again resembles abstract syntax:
-\begin{lstlisting}[numbers=none]
-fun example_fixed(name : string) => 
-  P.Seq(P.Str(name), P.Seq(pattern "SSTR: ESTR", ssn)) (* ssn as above *)
-\end{lstlisting}
+%The (unstated) intent here was to treat \lstinline{name} as a sub-pattern matching only itself, but this is not the observed behavior when \lstinline{name} contains special characters that have other meanings in patterns.
+
 Note that both functions have the same type and behave identically at many inputs, particularly those that would be expected during typical executions of the program (i.e.
-alphabetic names), so testing can miss the problem.
-Proving that such problems cannot arise involves reasoning about complex run-time data flows, so it is once again difficult to automate.
-In applications that query sensitive data, mistakes like this lead to \emph{injection attacks}, which are both common and catastrophic from the perspective of software security \cite{owasp2013}.
+alphabetic names).
+It is only when \lstinline{name} contains special characters that have meaning in the concrete syntax of patterns that a problem arises.
+In applications that query sensitive data, problems like this lead to \emph{injection attacks}, which are common and can be catastrophic from the perspective of software security \cite{owasp2013}.
+Proving that mistakes like these have not been made involves reasoning about complex run-time data flows, so it is once again difficult to automate.
  %Ultimately, of course, mistakes like this are the fault of a programmer using a flawed heuristic, and they could be avoided with discipline.
@@
 
-	\subsection{Existing Approaches}\label{sec:syntax-existing} The difficulties described above are not unique to regular expression patterns.
-Whenever a library  encourages the use of dynamic string parsing solely to address the issue of syntactic cost (a fundamentally static issue), such problems arise.
+The difficulties above are not unique to regular expression patterns.
+Whenever a library encourages the use of dynamic string parsing to address the issue of syntactic cost (which is, fundamentally, not a dynamic issue), such problems can arise.
 %Strings are, simply put, not ideally suited for this task.
 	This has motivated much research on reducing the need for dynamic string parsing \cite{Bravenboer:2007:PIA:1289971.1289975}.
-Existing approaches can be broadly classified as being based on either \emph{direct syntax extension} or \emph{static term rewriting}, described in Secs.
-\ref{sec:syntax-extension} and \ref{sec:term-rewriting}, respectively, below.% We summarize the most relevant work below.
+Existing alternatives can be broadly classified as being based on either \emph{direct syntax extension} or \emph{static term rewriting}.
+We describe these next, in Secs.
+\ref{sec:syntax-extension} and \ref{sec:term-rewriting} respectively.% We summarize the most relevant work below.
 
 \subsubsection{Direct Syntax Extension}\label{sec:syntax-extension}
-One approach we might consider is to use a system that allows us to directly extend the syntax of our language with new derived syntactic forms for regular expression patterns.% for patterns.
-
-The simplest such syntax extension systems are those where each new syntactic form is defined by a single rewrite rule.
+One tempting approach is to use a system that gives library providers the power to directly extend the concrete syntax of our language with new derived syntactic forms.
+%for regular expression patterns.% for patterns.
+
+The simplest such systems are those where the elaboration of each new syntactic form is defined by a single rewrite rule.
 For example, Gallina, the ``external language'' of the Coq proof assistant, supports such extensions \cite{Coq:manual}.
@@
 Unfortunately, these systems are not flexible enough to allow us to express pattern syntax in the conventional manner.
-For example, sequences of characters  can only be parsed as identifiers using these systems, rather than as characters in a pattern.
-
-Syntax extension systems based on context-free grammars like  Sugar* \cite{erdweg2013framework}, Camlp4 \cite{ocaml-manual} and many others are more expressive, and would allow us to directly introduce pattern syntax into our core language's grammar, perhaps like this:
+For example, sequences of characters  can only be parsed as identifiers using these systems, rather than as characters in a regular expression pattern.
+
+Syntax extension systems based on context-free grammars like  Sugar* \cite{erdweg2013framework}, Camlp4 \cite{ocaml-manual} and several others are more expressive, and would allow us to directly introduce pattern syntax into our core language's grammar, perhaps following Unix conventions like this:
 \begin{lstlisting}[numbers=none]
@@
 
-Unfortunately, this approach is rather perilous because none of the systems mentioned thusfar can guarantee that \emph{syntactic conflicts} between such extensions will not arise.
+Unfortunately, this approach is perilous because the systems mentioned thusfar cannot guarantee that \emph{syntactic conflicts} between such extensions will not arise.
 As stated directly in the  Coq manual: ``mixing different symbolic notations in [the] same text may cause serious parsing ambiguity''.
-Consequently, if another library provider used similar syntax for a different implementation or variant of regular expressions, or for some other unrelated abstraction, then a client could not simultaneously use both libraries in the same scope.
-Properly considered, every combination of extensions introduced using these mechanisms creates a \emph{de facto} derived syntactic dialect of our language.
+If another library provider used similar syntax for a different implementation or variant of regular expressions, or for some other unrelated construct, then a client could not simultaneously use both libraries in the same piece of code.
+So properly considered, every combination of extensions introduced using these mechanisms creates a \emph{de facto} derived syntactic dialect of our language.
 % Resolving such parsing amibiguities is left to each client of the library.
@@
 In response to this problem, Schwerdfeger and Van Wyk developed a modular analysis that accepts only grammar extensions that specify a universally unique starting token and obey subtle constraints on  the follow sets of base language non-terminals \cite{conf/pldi/SchwerdfegerW09}.
-Extensions that satisfy these criteria can be used together in any combination without the possibility of conflict.
+Extensions that satisfy these criteria can be used together in any combination without the possibility of syntactic conflict.
 However, note that the most natural starting tokens like \lstinline{pattern} cannot be guaranteed to be universally unique, so we would be forced to use a more verbose token like \lstinline{edu_cmu_verse_rx_pattern}.
@@
 
-Putting this aside, we must also decide how our newly introduced forms elaborate to forms in our base grammar.
-This is tricky because we have defined a modular encoding of patterns -- which particular module should the elaboration use?
-Clearly, simply assuming that some module identified as \lstinline{P} satisying \lstinline{PATTERN} is in scope is a brittle solution.
-In fact, we should expect that the extension mechanism actively prevents such capture of specific variable names to ensure that variables (including module variables) can be freely renamed.
-Such a \emph{hygiene discipline} is well-understood only when performing term-to-term rewriting (discussed below) or in simple rewrite systems like those found in Coq.
-For more flexible mechanisms, the issue is a topic of ongoing research (none of the grammar-based mechanisms described above enforce hygiene).
-
-Putting aside this question of hygiene as well, we can address the problem by requiring that the client explicitly identify the module the elaboration should use:
+Putting this aside, there is also the question of how newly introduced forms elaborate to forms in our base grammar.
+In our example, this is tricky because we have defined a modular encoding of patterns -- which particular module should the elaboration use?
+Clearly, simply assuming that some module identified as \lstinline{P} matching \lstinline{PATTERN} is in scope is a brittle solution.
+In fact, we should expect that the system actively prevents such capture of specific variable names to ensure that variables (including module variables) can be freely renamed.
+Such a \emph{hygiene discipline} is well-understood only when performing term-to-term rewriting (discussed below) or in simple language-integrated rewrite systems like those found in Coq.
+For mechanisms that operate strictly at the level of context-free grammars, this issue has not been addressed.
+
+Putting aside this question of hygiene as well, we can address the problem of choosing a module to use in the elaboration by requiring that the client explicitly identify it as an ``argument'' to the form:
 \begin{lstlisting}[numbers=none]
@@
 \end{lstlisting}
-For patterns constructed compositionally, we can define \emph{splicing} syntax for both strings and other patterns, e.g. rewriting \lstinline{example_bad} as shown below:
+For patterns constructed compositionally, we can define \emph{splicing} syntax for both strings (using prefix \lstinline{@}) and other patterns (using prefix \lstinline{%}).
+For example, we can write \lstinline{example_bad} instead like this:
 \begin{lstlisting}[numbers=none,escapechar=|]
-fun example_synexts(name : string) => 
+fun example_fixed1(name : string) => 
   edu_cmu_verse_rx_pattern P /@name: %ssn/
@@
 
-A general problem with approaches like these (in addition to those already mentioned) is that they  suffer from a paucity of direct reasoning principles -- given an unfamiliar piece of syntax, there is no simple method for determining what type it will have, or even for identifying which extension determines its elaboration, causing difficulties for both humans (related to code comprehension) and tools.
+A general problem with approaches like these (in addition to those already mentioned) is that they  suffer from a paucity of direct reasoning principles -- given an unfamiliar piece of syntax, there is no straightforward method for determining what type it will have, or even for identifying which extension determines its elaboration, causing difficulties for both humans (related to code comprehension) and tools.
 
@@
 An alternative approach is to leave the concrete syntax of the language fixed, but repurpose it for novel ends using a \emph{local term-rewriting system}.
-Among the earliest examples is the LISP macro system \cite{Hart63a}, later modified, notably in the Scheme language, to support reasoning hygienically about the rewriting that is performed \cite{Kohlbecker86a}.
-In languages with a stronger static type discipline, variations on macros that restrict rewriting to a particular type and perform the rewriting in an explicitly staged manner have also been studied \cite{Herman10:Theory,ganz2001macros} and integrated into languages, e.g.
+The LISP macro system \cite{Hart63a} is the most prominent example of such a system.
+Na\"ively, this system also suffers from the problem of unhygienic variable capture, but  modifications of this system, notably in the Scheme dialect of LISP, support reasoning hygienically about the rewriting that a macro performs \cite{Kohlbecker86a}.
+In languages with a stronger static type discipline, variations on macros that restrict rewriting to a particular type and perform the rewriting statically have also been studied \cite{Herman10:Theory,ganz2001macros} and integrated into languages, e.g.
 MacroML \cite{ganz2001macros} and Scala \cite{ScalaMacros2013}.
@@
 
-While this does not leave us with syntax that is quite as clean as would be possible with a  na\"ive syntax extension,  it does address many of the problems we discussed above: there cannot be syntactic conflicts (because the syntax is not extended), we can define macro abbreviations because macros are integrated into the language, there is a hygiene discipline that guarantees that the elaboration will not capture variables inadvertently, and by using a typed macro system, there is a typing discipline (i.e.
+While this does not leave us with syntax that is quite as clean as would be possible with a  na\"ive syntax extension,  it does avoid many of the problems with that approach: there cannot be syntactic conflicts (because the syntax is not extended), we can define macro abbreviations because macros are integrated into the language, there is a hygiene discipline that guarantees that the elaboration will not capture variables inadvertently, and by using a typed macro system, there is a typing discipline (i.e.
 we need not examine the elaboration directly to know what type it must have).
@@
 \subsection{Contributions}\label{sec:syntax-contributions}
-We propose a new primitive -- the \textbf{typed syntax macro} (TSM) -- that combines the flexibility of syntax extensions with the reasoning guarantees of statically-typed macros.
-This will address all of the problems brought up thusfar.
+We will now introduce a new primitive -- the \textbf{typed syntax macro} (TSM) -- that combines the syntactic flexibility of syntax extensions with the reasoning guarantees of statically-typed macros, and show how to integrate it with the Verse module system.
+This will address all of the problems brought up in the previous two subsections.
 We will introduce TSMs in Sec.
-\ref{sec:tsms} below.
+\ref{sec:tsms}.
 
 We then go further in Sec.
-\ref{sec:tsls}, proposing \textbf{type-specific languages} (TSLs).
+\ref{sec:tsls}, introducing \textbf{type-specific languages} (TSLs).
 TSLs are TSMs associated directly with a type when it is generated.
-Verse then leverages local type inference to implicitly control TSL dispatch, further decreasing syntactic cost (which is, of course, the entire purpose of this exercise).
-
-The result is that derived syntax defined by library providers using TSMs and TSLs provides client with a syntactic cost cost comparable to that of derived syntax built in primitively by a language designer or  using a na\"ive tool like Camlp4, obviating the need for these approaches.
+Verse leverages local type inference to implicitly control TSL dispatch, further decreasing syntactic cost (which is, of course, the entire purpose of this exercise).
+
+The result is that derived syntax defined by library providers using TSMs and TSLs provides comes at syntactic cost comparable to that of derived syntax built in primitively by a language designer or  using a na\"ive tool like Camlp4, without the associated problems.
 
@@
 \end{lstlisting}
-Here, a \emph{parameterized TSM}, \lstinline{pattern}, is being applied first to a module parameter, \lstinline{P}, then to a {delimited form}, \lstinline{/SURLA|T|G|CEURL/} (a number of alternative delimiters are also provided by Verse's concrete syntax and could equivalently be used).
+Here, a \emph{parameterized TSM}, \lstinline{pattern}, is being applied first to a module parameter, \lstinline{P}, then to a \emph{delimited form}, \lstinline{/SURLA|T|G|CEURL/} (a number of alternative delimiters are also provided by Verse's concrete syntax and could equivalently be used).
 The TSM statically parses the \emph{body} of the provided delimited form, i.e.
@@
 another external expression.
-In this case, the term above elaborates to the following term (written concretely):
+In this case, \lstinline{pattern} generates the following elaboration (written concretely):
 
@@
 \end{lstlisting}
-Note that identifying the module parameter as \lstinline{P} here is not important, i.e.
+We first identify the TSM, then specify the module parameter.
+Note that identifying it as \lstinline{P} here is not important, i.e.
 the TSM can be used with \emph{any} module matching signature \lstinline{PATTERN}.
@@
 
-A hygiene mechanism ensures that only portions of the generated elaboration derived from such spliced expressions can refer to the variables at the use site, preventing inadvertent variable apture by the elaboration.
+A hygiene mechanism ensures that only portions of the generated elaboration derived from such spliced expressions can refer to the variables at the use site, preventing inadvertent variable capture by the elaboration.
 Put another way, the  elaboration logic must be valid in any typing context.
@@
 We have described and given a more detailed formal specification of TSMs in a recently published paper \cite{sac15}, and TSLs in a paper last year \cite{TSLs}, both in the context of the Wyvern language.
-The mechanics of elaboration encoding/decoding and hygienic elaboration validation have all been satisfyingly specified in this context.
-
-Wyvern does not specify an ML-style module system (type declarations provide the only form of type generativity), so in the context of Verse, there are some extensions that need to be defined in the formalization to handle module parameters, as mentioned earlier.
-We plan to complete this in the course of writing the dissertation, and do not anticipate it requiring more than 2-3 weeks of effort over Summer 2015.
-We emphasize that we do not plan to specify the full module system in detail; we will assume a standard specification of the module system and specify only those portions that interface with the novel aspects of the core language.
+The basic mechanics of elaboration encoding/decoding and hygienic elaboration validation have all been satisfyingly specified in this context.
+
+Wyvern does not specify an ML-style module system (type declarations provide the only form of type generativity), so in the context of Verse, there are some extensions that need to be defined in the formalization to handle module parameters.
+We plan to complete this in the course of writing the dissertation, and anticipate it requiring about 3 weeks of effort over Summer 2015.
+We emphasize again that we do not plan to specify the full module system in detail; we will assume that judgements of an appropriate form have been specified and detail only those portions that interface with the novel aspects of the core language.
 
@@
 \section{Modularly Metaprogrammable Type Structure}\label{sec:metamodules}
-Let us now turn our attention to the type structure of the Verse external language (EL).
+Let us now turn our attention to the type structure of the Verse EL.
 In contrast to other languages, which build in constructs like case types (shown briefly in use in the previous section), tuples, records, objects and so on primitively, in Verse these can all be expressed as modes of use of \emph{metamodules} (which are distinct from, though conceptually related to, {modules}, as we will see).
@@
 \subsection{Motivating Example: Labeled Tuples and Regular Strings}\label{sec:metamodules-example}
-As a simple introductory example, let us define a type classifying conference papers:
+As a simple introductory example, let us consider a type classifying conference papers:
 \begin{lstlisting}[numbers=none]
@@
 
-Our take on labeled tuples supports a number of other basic operations.
+Labeled tuples support a number of other basic operations.
 For example, labeled tuples can be concatenated (with any common components updated with the value on the right) using the \lstinline{ltuple+} operator:
 \begin{lstlisting}[numbers=none]
-let val a : ltuple {SPCTauthors : EPCTlist(rstring /SURL.+EURL/)} = {["SUSHarry Q. BovikEUS"]}
+let val a : ltuplex {SPCTauthors : EPCTlist(rstring /SURL.+EURL/)} = {["SUSHarry Q. BovikEUS"]}
 let val exmpl_paper_final = ltuple+ exmpl_paper a
@@
 
-If on the other hand we do care about preserving the type disequalities in question, the workaround is even less elegant: we first need to define a record type with component labels tagged in some sufficiently unambiguous way by position, e.g.:
+If on the other hand we do care about preserving the type disequalities in question, the workaround is even less elegant.
+We first need to define a record type with component labels tagged in some sufficiently unambiguous way by position, e.g.:
 \begin{lstlisting}[numbers=none]
@@
 \end{lstlisting}
-Ensuring that there will not be an out-of-bounds error here would again require extrinsically reasoning about the term.
+Ensuring that there will not be an out-of-bounds error here would again require more sophisticated extrinsic reasoning.
 %Refinement types are strictly a layer above the language, and thus cannot affect its dynamic semantics.
@@
 As a result, a compiler cannot use the invariants they capture to optimize the representation of a value.
-For example, the fact that some value of type \lstinline{string} can be given the refined type \lstinline{rstring /A|B/} cannot be justification to locally alter its representation to, for example, a single bit, both because it could later be passed into a function expecting a standard string and because there are many other possible refinements.
+For example, the fact that some value of type \lstinline{string} can be given the refinement \lstinline{rstring /A|B/} cannot be justification to locally alter its representation to, for example, a single bit, both because it could later be passed into a function expecting a standard string and because there are many other possible refinements.
 Of course, for regular strings, there may be limited benefit (or even some cost) to controlling representation differentially, but in other situations, this could have significant performance ramifications.
@@
 %Object systems are perhaps even more diverse across languages\todo{citation}.
-%Other dialects are justified by a desire to provide more special-purpose semantic constructs. For example, 
+%Other dialects are justified by a desire to proivde more special-purpose semantic constructs. For example, 
 %Substantially more sophisticated examples require at most only a very slightly more capable intermediate language (the IL of popular virtual machines, even those with flaws like ubiquitous \texttt{null} references, e.g.
@@
 \end{itemize}
+
+Static values are values of the \emph{typed static language} (SL), which itself forms a total typed lambda calculus, with \emph{static expressions} $\sigma$ classified by \emph{kinds} $\kappa$.
+Types are one kind of static value.
+We discuss other kinds of static values below.
+
 \subsubsection{Example: Labeled Tuple Types via Metamodules}
-\begin{figure}[hp!]
+\begin{figure}[p!]
 \begin{lstlisting}
@@
 
-syntax ltplx(L :: LTUPLEX, param :: LblTyOrdMap) at L.c(param) {
+^\label{line:ltplx-tsm-start}^syntax ltplx(L :: LTUPLEX, param :: LblTyOrdMap) at L.c(param) {
   fn ps => (* 
@@
   *)
-}
+}^\label{ltplx-tsm-end}^
 
@@
   tycon c {
-  	(* type translation schema: *)
+  	(* type translation generator: *)
   	fn (param :: LblTyOrdMap) => fold (list_of_ltom param)
@@
 
-  syn opcon intro_labeled {
+  ^\label{line:introl-start}^syn opcon intro_labeled {
   	fn (param :: list(Lbl), args :: list(Arg)) => 
       (* ... same as intro_unlabeled but reorder first ... *)
-  }
-
-  syn opcon # {
+  }^\label{line:introl-end}^
+
+  ^\label{line:prj-start}^syn opcon # {
   	fn (param :: Lbl, args :: list(Arg)) =>  
       (* ... generate the appropriate n-fold projection ... *)
-  }
-
-  syn opcon + {
+  }^\label{line:prj-end}^
+
+  ^\label{line:ltpl-conc-start}^syn opcon + {
   	fn (param :: unit, args :: list(Arg)) => 
       (* generate the new type and combine the two nested tuples *)
-  }
-
-  syn opcon - {fn (param : Lbl, args : list(Arg)) => 
+  }^\label{line:ltpl-conc-end}^
+
+  ^\label{line:ltpl-drop-start}^syn opcon - {fn (param : Lbl, args : list(Arg)) => 
     (* generate the new type and drop the appropriate element *)
-  }
-} with syntax ltplx for c 
+  }^\label{line:ltpl-drop-end}^
+} with syntax ltplx for c ^\label{line:ltplx-annotation}^
 (* synonyms used in Sec. 5.1 *)
@@
 The metamodule \lstinline{Ltuplex} defines a \emph{type translation schema}, a static function that determines the translations of types constructed by \lstinline{c} (each identified by the value of its parameter).
-Internal types are encoded as static values of type \lstinline{ITy}, which we assume has a TSL associated with it that allows us to use standard \emph{quasiquotation} syntax.
+Internal types are encoded as static values of kind \lstinline{ITy}, which we assume supports standard \emph{quasiquotation} syntax.
 Here, the  type translation schema on lines \ref{line:c-start} to \ref{line:c-end} generates nested internal products by folding over a list generated from the type parameter (in the standard way), returning the encoding of \lstinline{unit} for the empty case, and recursively nesting encodings of binary products otherwise using the standard \emph{unquote} operator, \lstinline{%}.
@@
 
-If we assume that all regular string types translate to standard internal strings, then the translation of \lstinline{Paper} is ultimately:
+If, for example, all regular string types translate to standard internal strings, encoded by some internal type abbreviated \lstinline{string}, and substitute this in, then the translation of \lstinline{Paper} is ultimately:
 \begin{lstlisting}[numbers=none]
@@
 
-\paragraph{Operator Constructors} On line \ref{line:intro-unlabeled-sig} of Figure \ref{fig:ltuplex}, the metasignature \lstinline{LTUPLEX} specifies an operator constructor, \lstinline{intro_unlabeled} parameterized by static values of kind \lstinline{unit}.
+\paragraph{Operator Constructors} On line \ref{line:intro-unlabeled-sig} of Figure \ref{fig:ltuplex}, the metasignature \lstinline{LTUPLEX} specifies an operator constructor, \lstinline{intro_unlabeled}, parameterized by static values of kind \lstinline{unit}.
 The qualifier \lstinline{ana} specifies that this opcon can only be used in analytic positions (i.e.
@@
 The function must return an encoding of an  internal expression, which will become the translation of the external expression.
-Internal expressions are encoded as static values of kind \lstinline{IExp}, and again we assume a TSL has been defined so that we can use standard quasiquote syntax.
-
+Internal expressions are encoded as static values of kind \lstinline{IExp}, and again we assume that we can use standard quasiquote syntax.
+Here, we simply generate the trivial value for an empty labeled tuple and recursively generate nested pairs otherwise, following directly the structure for the type translation generator above.
+This is, of course, intentional: the introductory operation for a type must generate an expression translation consistent with the corresponding type translation for type safety to hold.
+
+\todo{describe ltplx?}
+
+ The remaining opcons are synthetic opcons, which means they can be used anywhere.
+They operate similarly, but rather than taking a type as input, they produce a pair of a type and a translation.
+For concision, we omit the details.
+Instead, note that the code written in these definitions is essentially exactly what would appear in a standard implementation of the first stage of a compiler.
+The novelty is not in the definitions, but in how they are organized and brought into the language.
 \subsection{Timeline}\label{sec:metamodules-timeline}
+For the sake of concision, we leave providing substantially more technical details on how metamodules operate as work that remains to be done.
+However, we have a lengthy draft of a technical report that provides these details (for an older version of this work).
+We anticipate completing this work will take about 2 months during Summer 2015.
+
 % The semantics uses internal type abstraction to guarantee that metamodule providers can reason modularly about translation invariants, just as it uses external type abstraction to guarantee that module providers can reason modularly about representation invariants.%For example, a metamodule that expresses the type structure of records wouldLblTyList introduce a type constructor  \texttt{record} parameterized by a static mapping from labels to types, and two operators: the introductory operator \texttt{intro}, parameterized by a list of row labels, and the row projection operator \texttt{prj}, parameterized by a row label.

### 2015-05-28 atlam-icfp15/atlam-icfp15.tex
ICFP 15

--- 2014-12-02 att-pldi15/att-pldi15.tex
@@
 %\documentclass[12pt]{article}
-\documentclass[pldi]{sigplanconf-pldi15}
+\documentclass[preprint]{sigplanconf}
 % The following \documentclass options may be useful:
@@
 
+\newcommand{\Verse}{{\textsf{\small Verse}}}
+
+\definecolor{mygray}{rgb}{0.5,0.5,0.5}
 \usepackage{times}
@@
 basicstyle=\ttfamily\footnotesize,
-morekeywords={newcase,extends}}
+morekeywords={module,tycon,tcsig,translation,ana,syn,intro,indexed,by,tcstruct},
+literate={~} {$\sim$}{1},
+numbersep=5pt,
+numberstyle=\small\ttfamily\color{mygray}}
 
@@
 
-\newcommand{\moutput}{^{\color{gray}+}}
+\newcommand{\moutput}{^{\color{gray}-}}
 \newcommand{\rulename}[1]{(#1)}
@@
 \usepackage[compact]{titlesec}  
-\titlespacing{\section}{0pt}{6pt}{2pt}
-\titlespacing{\subsection}{0pt}{3pt}{0pt}
-\titlespacing{\subsubsection}{0pt}{3pt}{0pt}
-\titlespacing{\paragraph}{0pt}{3pt}{5pt}
+\titlespacing*{\section}{0pt}{0pt}{3pt}
+\titlespacing*{\subsection}{0pt}{3pt}{0pt}
+\titlespacing*{\subsubsection}{0pt}{3pt}{0pt}
+\titlespacing*{\paragraph}{0pt}{3pt}{5pt}
 
@@
 % known bug: http://tex.stackexchange.com/questions/1522/pdfendlink-ended-up-in-different-nesting-level-than-pdfstartlink
+
+\newcommand{\liv}[1]{\lstinline{#1}}
+\usepackage{enumitem}
 
@@
 
-\title{Modular Composing Typed Language Fragments}
+\title{Modularly Metaprogrammable Type Structure}
 %\subtitle{Modular Type Constructors}
 
-\authorinfo{}{}{}
+\authorinfo{\vspace{-10px}}{}{}
 %\authorinfo{Cyrus Omar \and Jonathan Aldrich}
@@
 \begin{abstract}
-Researchers often describe type systems as fragments, leaving to language designers the task of composing these to form complete programming  languages. 
-This has not been a systematic process: metatheoretic results must be established anew for each composition of fragments, guided only informally by metatheorems derived for simpler systems.
-As the language design space grows, mechanisms that provide stronger modular reasoning principles than this are needed.
-
-In this paper, we take a foundational approach, specifying an extensible typed translation semantics, @$\lambda$.
-Only the $\rightarrow$ type constructor is built in; all others (we discuss constrained strings and variants of record types) are defined by extending a \emph{tycon context}.
-Each tycon defines associated term-level \emph{opcons} (e.g.
-row projection) using a static language where types and translations are values.
-The semantics come with modular metatheoretic guarantees, notably \emph{type safety} and \emph{conservativity}: that \emph{tycon-specific  invariants} established in any ``closed world'' are conserved in the ``open world''.
-Remarkably, extension providers do not need to  provide the semantics with mechanized specifications or proofs.
-Instead, these guarantees arise from a form of translation validation that, taking inspiration from the ML module system, uses type abstraction to check that the opcons associated with a tycon respect the \emph{translation independence} of all others.
+% 
+% This has not been a systematic process: metatheoretic results must be reestablished for each combination of fragments, guided only informally by metatheorems derived for simpler systems.
+% As the language design space grows, mechanisms that systematize this process and provide stronger modular reasoning principles than this are needed.
+
+We introduce \Verse, a programming language specified, like  many modern languages are, by a {typed translation semantics} targeting\- a common intermediate language (e.g.
+CIL bytecode, though for simplicity, we use a simply typed lambda calculus).
+In contrast to large such languages like Scala, the external language of {\Verse} is remarkably small, building in only function types directly.
+In lieu of other type structure, {\Verse} provides a metaprogramming mechanism that gives library providers the ability to modularly express type structure of their own design.
+For example, record types are not built in, but they can be  expressed by defining a \emph{tycon structure} that introduces a type constructor, \texttt{record},
+%indexed by a finite mapping from static labels to types, 
+together with record introduction and row projection operators.
+The logic governing the typechecking and translation of these operators is defined using a static metalanguage where types and translations are manipulated as values.
+%In other words, the mapping from the external language to the internal language is governed by metaprograms written in the static language, rather than fixed \emph{a priori}.
+
+{\Verse} performs translation validation to guarantee type safety and \emph{conservativity}: that a broad class of metatheorems about such logic, including those establishing {translation invariants}, need only be established in a ``closed world'', i.e.
+as if the collection of tycon  structures was finite.
+These are necessarily conserved in the ``open world'', i.e.
+when new tycon structure are introduced, with no new proof obligations.
+This critical modularity result relies on type abstraction to ensure that each tycon structure maintains \emph{translation independence} with respect to all others.
 \end{abstract}
@@
 %extensible languages; module systems; type abstraction; typed compilation; type-level computation
+
 \section{Introduction}\label{intro}
-Typed programming languages are usually described as consisting of \emph{fragments}, each contributing separately to the concrete and abstract syntax and static and dynamic semantics of the language \cite{pfpl}.
-Harper organizes such fragments around \emph{type constructors}, e.g.
-$\rightarrow$ for total function types, describing each in a different chapter of his book.
-Languages are then identified by a set of type constructors, e.g.
-$\mathcal{L}\{{\rightharpoonup}\,{\forall}\,{\mu}\,{\keyw{1}}\,{\times}\,{+}\}$ is the language of partial (i.e.
-possibly non-terminating) function types, polymorphic types, recursive types, nullary and binary product types and binary sum types (its syntax is shown in Figure \ref{syntax-IL}, discussed below).
-Another common practice is to describe a fragment using a simple calculus having a ``catch-all'' constant and base type to stand notionally for all ``other'' terms and types.% (e.g.
-\cite{sanitation-psp14}).
-
-In contrast, the usual metatheoretic reasoning techniques for programming languages  (e.g., rule induction) operate on complete language specifications.
-Each {combination} of fragments must formally be treated as its own monolithic language for which  metatheorems must be established anew, guided only informally by those derived for the smaller systems from which the language is notionally composed.
-
-%In both cases, it is left as an extrinsic concern to ensure that the metatheory developed separately is actually conserved when fragments defined in these ways are composed to form a complete language.
-
-This is not an everyday problem for programmers only because fragments like those mentioned above are ``general purpose'': they make it possible to \emph{isomorphically embed}  many other fragments as ``libraries''.
+
+A key feature of the ML {module system} is that it supports type abstraction at interface boundaries \cite{MacQueen:1984:MSM:800055.802036}.
+This enables programming ``in the large'' by localizing reasoning about data representation invariants and making it possible to  evolve these representations without breaking client programs.
+%By holding the representation of a type abstract outside of a module, the language guarantees that representation invariants at that type need only be established locally, in the ``closed world'' of the module.
+Introducing a new module into a program does not  require reestablishing such invariants.
+There is substantial agreement between dialects of ML about the value of this fundamental feature.
+Other languages are also increasingly converging on an ML-like design, e.g.
+Haskell \cite{Kilpatrick:2014:BRH} and Scala \cite{conf/oopsla/AminRO14}.
+\Verse, the language that we will introduce in this paper, does not buck this trend, providing a module system essentially identical to the Standard ML module system \cite{MacQueen:1984:MSM:800055.802036}.
+%An example is discussed in the supplement.%We make only syntactic changes, discussed in the supplement, consistent with Verse's layout-sensitive concrete syntax.
+
+Where there are more substantial differences between languages is with regard to the semantics of the underlying core language, i.e.
+the language of types and expressions.
+Indeed, the language design community has produced hundreds of full-scale languages, toy dialects and minimal calculi that explore different points in the design space of core language constructs.
+For example:
+%\vspace{-4px}
+\begin{itemize}
+\setlength{\itemsep}{-1pt}
+\item 
+\textbf{General Purpose Constructs:} 
+Many variations on products exist: 
+$n$-ary tuples, 
+labeled tuples, 
+records (identified up to reordering), records with extension and update operators\footnote{The Haskell wiki states that ''extensible records are not implemented in GHC.
+The problem is that the record design space is large, and seems to lack local optima.
+[...] As a result, nothing much happens.'' \cite{GHCFAQ}} \cite{ocaml-manual}, and
+records with width and depth coercions \cite{Cardelli:1984:SMI:1096.1098}%, 
+ %mutable fields \cite{ocaml-manual}, 
+%field delegation \cite{atlang-gpce14} \todo{gpce submission} 
+%and 
+% ``methods'' (i.e. pure objects) \cite{TSLs}
+. 
+Disjunctive types also come in variants: standard datatypes, 
+open datatypes \cite{conf/ppdp/LohH06,journals/toplas/MillsteinBC04}, 
+polymorphic variants \cite{ocaml-manual} and 
+exception types \cite{Tofte:89:TheDefinitionOfStandardML}. Combinations occur as class-based object systems, of which there are a seemingly endless variety.
+
+\item
+\textbf{Specialized Constructs:} More specialized constructs are also commonly introduced in dialects, e.g.
+for distributed programming \cite{Murphy:2007:TDP:1793574.1793585}, concurrency \cite{AliceLookingGlass}, reactive programming \cite{mandel2005reactiveml}, databases \cite{Ohori:2011:MSM:2034773.2034815},  units of measure \cite{conf/cefp/Kennedy09}, typechecking XML schemas \cite{HosoyaVouillonPierce2000ICFP}, web programming \cite{conf/popl/Chlipala15} and string sanitation \cite{sanitation-psp14}.%\verb|sprintf| in the OCaml dialect statically distinguishes format strings from strings.% All of these are implemented as dialects of existing languages, presumably because a strong encoding was not feasible.
+
+\item
+l\textbf{Foreign Constructs:} A safe and natural foreign function interface (FFI) can be valuable for interacting with legacy or low-level device code.
+This requires enforcing the type system of the foreign language in the calling language.
+%Using  a FFI that does not do this can lead to safety issues, even when both languages are separately known to be safe.
+%Safe FFIs generally require direct extensions to the language. 
+For example, MLj builds in a safe FFI to Java \cite{Benton:1999:IWW:317636.317791}.
+\end{itemize}
+\vspace{-3px}
+
+% Such problems  do not arise for constructs that are expressed inside a language with a modern \emph{module system}, like ML, because the module system .
+For example, a module for sets in ML can hold its representation type abstract, ensuring that any invariants necessary to maintain the isomorphism need only be maintained by the functions in the module (e.g.
+uniqueness, if using a list representation).
+These then continue to hold no matter which other modules are in use by a client \cite{harper1997programming}.% Other languages expose \emph{abstract data types} in other ways, also to localize reasoning \cite{liskov1974programming}.
+%Mechanisms that can help decrease the complexity of an embedding without violating abstraction barriers are thus valuable, and we will lead into our work in Sec.
+\ref{prior-work} by summarizing them.
+
+% %In both cases, it is left as an extrinsic concern to ensure that the metatheory developed separately is actually conserved when fragments defined in these ways are composed to form a complete language.
+
+% This is not an everyday problem for programmers only because fragments like those mentioned above are ``general purpose'': they make it possible to \emph{isomorphically embed}  many other fragments as ``libraries''.
 For example, a list type constructor need not be built in because lists are isomorphic to the type $\iforall{\alpha}{\imu{t}{\iunit + (\alpha \times t))}}$ (datatypes in ML combine these into a single declaration construct).
@@
 %But while universality properties (e.g. ``Turing completeness'') are often enough to guarantee that an embedding that preserves a desirable fragment's dynamic semantics can be constructed, 
-Establishing an isomorphic embedding of a fragment is not always possible, however, because it requires preserving both the static and dynamic semantics and, if defined, performance bounds specified by a cost semantics, of all operators.
+% Establishing an isomorphic embedding of a fragment is not always possible, however, because it requires preserving both the static and dynamic semantics and, if defined, performance bounds specified by a cost semantics, of all operators.
 This is possible if they have ``function-like'' semantics, as the list operators do.
@@
 Each time such a situation occurs,  a new \emph{dialect} is needed.
-The ML lineage, like others, faces a proliferation of dialects that go beyond ``core ML'':%, showing that this is not an isolated example:
+%, showing that this is not an isolated example:
 %Reynolds, in a remark that recalls the ``Turing tarpit'' of Perlis \cite{Perl82a}, summarizes the issue \cite{Reynolds94anintroduction}: 
@@
 
-\begin{enumerate}
-\vspace{-5px}
-%\compresslist
-\item 
-\textbf{General Purpose Fragments:} 
-Many variations on product types, for example, exist: 
-$n$-ary tuples, 
-labeled tuples, 
-records (identified up to reordering), and 
-records with width and depth subtyping \cite{Cardelli:1984:SMI:1096.1098} and functional update operators\footnote{The Haskell wiki notes that ''No, extensible records are not implemented in GHC.
-The problem is that the record design space is large, and seems to lack local optima.
-[...] As a result, nothing much happens.'' \cite{GHCFAQ}} \cite{ocaml-manual}%,
- %mutable fields \cite{ocaml-manual}, 
-%field delegation \cite{atlang-gpce14} \todo{gpce submission} 
-%and 
-% ``methods'' (i.e. pure objects) \cite{TSLs}
-. 
- Sum-like types are also exposed variously: standard datatypes, 
-open datatypes \cite{conf/ppdp/LohH06,journals/toplas/MillsteinBC04}, 
-polymorphic variants \cite{ocaml-manual} and 
-exception types \cite{Tofte:89:TheDefinitionOfStandardML}. Combinations occur as class-based object systems \cite{ocaml-manual}.
-
-\item
-\textbf{Specialized Fragments:} Fragments providing more specialized operators are also commonly introduced in dialects, e.g.
-for distributed programming \cite{Murphy:2007:TDP:1793574.1793585}, concurrency \cite{AliceLookingGlass}, reactive programming \cite{mandel2005reactiveml}, databases \cite{Ohori:2011:MSM:2034773.2034815},  units of measure \cite{conf/cefp/Kennedy09} and regular string sanitation \cite{sanitation-psp14}.%\verb|sprintf| in the OCaml dialect statically distinguishes format strings from strings.% All of these are implemented as dialects of existing languages, presumably because a strong encoding was not feasible.
-\vspace{-3px}
-\item
-\textbf{Foreign Fragments:} A safe and natural foreign function interface (FFI) can be valuable.
-This requires enforcing the type system of the foreign language in the calling language.
-%Using  a FFI that does not do this can lead to safety issues, even when both languages are separately known to be safe.
-%Safe FFIs generally require direct extensions to the language. 
-For example, MLj builds in a safe FFI to Java \cite{Benton:1999:IWW:317636.317791}.
-\end{enumerate}
-\vspace{-5px}
-
-This \emph{dialect-oriented} state of affairs is unsatisfying.
+This \emph{dialect-oriented} state of affairs is profoundly anti-modular.
 %Language designers are burdened with needing to understand the complete metatheory of every fragment they add to their language to make sure it is not incompatible with existing fragments, so languages change rarely.
@@
 many academic researchers) by making them unavailable to programmers.
-While programmers can choose from dialects supporting, e.g., a principled approach to distributed programming, or one that builds in support for statically reasoning about units of measure, one that supports both fragments may not be available.
+In other words, there is generally no good way to interface with a library written using a dialect different from one's own when a direct FFI is not available.
+There are, of course, bad ways to attempt  this.
+For example, one can find a compiler for the foreign dialect that targets a language for which an FFI is available, e.g.
+JVM or CIL bytecode, and then program against the code it generates.
+But this bypasses the language semantics (i.e.
+the interface), exposing implementation details directly.
+Only the type system of the intermediate language, not of the dialect itself, constrains this action.
+If the compiler maintains any stronger translation invariants, there is  no guarantee that they will be maintained.
+Moreover, if a different compiler is used, or the  compiler evolves its choices regarding data representations, client code can break.
+%Simply put, interacting directly with code a compiler generates is unlikely to be a healthy experience.
+
+It is clear that the only scalable way to program ``in the large'' is to stick to a single programming language that provides strong modular reasoning principles.
+Once one accepts that dialects are better rhetorical vehicles than practical artifacts, there are two options.
+One, exemplified by languages like Scala and Ocaml, is to attempt to build in a large number of constructs into the core language \emph{a priori}, and progressively add new ones as they emerge in backwards compatible revisions.
+Power is centralized around a small group of language designers.
+This results in a large core language, but because of a ``long tail'' effect, constructs that are situationally useful, or where there is  disagreement, can still end up marginalized.
+Other constructs that are included but turn out to be  flawed remain entrenched (and monopolize finite ``syntactic resources'', as we will discuss below).
+
+{\Verse} aims to explore the polar opposite end of this spectrum: a core language that builds in only minimal type structure \emph{a priori}.
+Instead, the core language is organized around a metaprogramming mechanism that gives library providers the ability to introduce new type structure of their own design, within reasonable constraints.
+
+The main challenge in decentralizing the semantics of a core language  comes in maintaining  strong metatheoretic properties.
+For example, if we simply conceived of the core language's semantics as being specified by a ``bag of rules'' and let library providers introduce new rules, one could easily define rules that violated type safety, interfered with invariants being maintained by rules defined in other libraries, or introduced non-determinism.
+There is also a question of how to allocate finite syntactic resources.
+For example, as the footnote above noted, the design space around record types is large.
+If we allow libraries to explore this design space, which library gets to determine the meaning of a form like \liv{\{label1: e1, label2: e2\}}?
+
+This paper is intended to introduce principled solutions to these problems.
+The key construct that the core language is organized around is called the \emph{tycon structure}.
+A tycon structures defines a type constructor and the semantics of the operators associated with it using statically evaluated metafunctions.
+The language enforces an abstraction barrier around each tycon structure that ensures that they cannot interfere with one another, which preserves modular reasoning.
+Encouragingly, type abstraction is critical to this abstraction barrier, much like it is to the abstraction barrier between modules.
+A bidirectional type system is used to resolve ambiguities with regard to common forms, like the record introduction form shown above (following recent work on ambiguity-free syntax extensions in the Wyvern programming language \cite{TSLs}).
+
+We will discuss labeled tuples with functional update, records and \emph{regular strings} (based on a recent specification \cite{sanitation-psp14}) as examples of constructs that are, or would need to be, built in to comparable languages but that can be expressed as modular tycon structures in {\Verse}.
+This mechanism is not all-encompassing -- there are many compelling constructs that cannot be expressed in {\Verse} today.
+We will discuss its limitations more specifically in Sec.
+\ref{sec:discussion}.
+But this paper sets the ``ground rules'' of the game: we expect future versions of the language, and competing languages, to provide metatheoretic guarantees at least as strong as those we establish here, while being ever more expressive.
+%% This paper is intended to introduce a novel mechanism that we argue puts it uniquely on a path toward this goal.
+
+The remainder of the paper is organized as follows:\vspace{-3px}
+\begin{itemize}[noitemsep]
+\item In Sec. 2, we give a tutorial-style introduction to \Verse. %We give intuition about the translation validation process that guarantees type safety and modular reasoning. 
+\item In Secs.
+3-4, we give a formal account of tycon structures and the translation validation process with a typed lambda calculus, $\lambda_\textsf{Verse}$.
+The accompanying supplement contains more details.
+\item In Sec. 5, we summarize the key metatheoretic properties of the calculus introduced in Sec. 3. 
+\item In Sec. 6, we compare tycon structures to other related work, and discuss present limitations and avenues for future work.
+\end{itemize}
+
+\section{\textsf{Verse} by Example}
+From the perspective of an everyday programmer, {\Verse} handles very much like other typed function programming languages.
+For example, let us build a (toy) academic conference management application in \Verse, beginning with some type synonyms:
+
+\vspace{-3px}\begin{lstlisting}
+let type Title = rstr /.+/
+let type Conf = rstr /([A-Z]+) (\d\d\d\d)/
+let type Paper = record {
+    title : Title,
+    conf : Conf
+}
+\end{lstlisting}\vspace{-3px}
+Starting from the bottom up, \liv{Paper} is a \emph{record type} that specifies two rows.
+The first is labeled \liv{title} and specifies type \liv{Title}, and the second is labeled \liv{conf} and specifies type \liv{Conf}.
+
+The types \liv{Title} and \liv{Conf} are \emph{regular string types}, which behave as strings known statically to be in a regular language specified by a regular expression \cite{sanitation-psp14}.
+Here, \liv{Title} specifies non-empty strings, and \liv{Conf} specifies conference names in the standard format, i.e.
+a series of capital letters followed by a four-digit year.
+Parentheses indicate captured groups.
+
+All types in {\Verse} arise by applying a \emph{type constructor} (\emph{tycon}, for short) to a statically valued \emph{type index}.
+Here, the tycon \liv{rstr} is indexed by statically valued regular expressions, and \liv{record} is indexed by statically valued finite mappings from labels to types.
+We are able to write the indices above using conventional concrete syntax not because this syntax is built in to {\Verse}, but because {\Verse} supports \emph{type specific languages} (TSLs), a feature recently introduced in the Wyvern programming language that permits modular reasoning about syntax extensions \cite{TSLs}.
+For concision, we will not go into a detailed discussion about TSLs in this paper.
+
+We can now define a function that generates papers as follows:
+\begin{lstlisting}
+fun exmpl_paper(title : Title) -> Paper
+  {title=title, conf="EXMPL 2015"}
+\end{lstlisting}
+and then call it and perform some further operations:
+\begin{lstlisting}
+let paper = exmpl_paper "Collapsing The Multiverse"
+let year = paper#conf#2
+\end{lstlisting}
+The value of \liv{year} projects out the row labeled \liv{conf} from the record \liv{paper}, and then the substring corresponding to the second captured group, i.e.
+the year.
+Had we specified an incorrect row label, or an out-of-bounds captured group index, we would encounter a static type error.
+Note also that type annotations were not needed on these bindings, because {\Verse} supports local type inference \cite{Pierce:2000:LTI:345099.345100}.
+The type of \liv{year} here is synthesized to be \liv{rstr /\\d\\d\\d\\d/}.
+
+If {\Verse} were organized like a standard language and simply built in the type structure that we saw being used above -- functions, records and regular strings -- this would be only somewhat novel (regular strings are not widely available in languages like Ocaml and Scala).
+But neither records nor regular strings are built directly in to {\Verse}.
+Instead, the core language of {\Verse} is structured like the first stage of a type-directed compiler, e.g.
+the TIL compiler for Standard ML \cite{tarditi+:til-OLD}.
+It consists of an \emph{external language} (EL) defined by a type-directed translation to a much simpler \emph{typed internal language} (IL).
+In other words, all external types and expressions (like those just defined) map, as they are being typechecked, to internal types and expression, respectively.
+Only the IL has a conventional dynamic semantics.
+
+Only function types are built in to the EL.
+The {\Verse} IL, for the purposes of this paper, builds in only functions, nullary and binary products, binary sums, recursive types, strings and numbers, i.e.
+it is reminiscent of a simple first-stage compiler IL.
+In practice, we would include a few other typical constructs, e.g.
+reference cells, contiguous arrays and concurrency primitives, but our design is largely independent of the details of the IL, so we omit discussion of these details.
+
+Ultimately, the translation generated for the example above will be the following (assuming suitable helper functions, not shown):
+\vspace{-3px}\begin{lstlisting}
+let type Title = string
+let type Conf = string
+let type Paper = Title * (Conf * unit)
+
+fun exmpl_paper(title : Title) -> Paper
+  (title, ("EXMPL 2015", ()))
+
+let paper = exmpl_paper "Collapsing the Multiverse"
+let year = groupPrj 2 /([A-Z]+) (\d\d\d\d)/ (
+  fst(snd(paper)))
+\end{lstlisting}\vspace{-3px}
+External functions  translate directly to internal functions.
+All other types and operations, however, delegate to \emph{statically evaluated metafunctions} associated with \emph{tycon structures}.
+Let us consider the \liv{record} tycon structure in Figure \ref{fig:record-conc}.
+
+The first line names the tycon structure and then ascribes a \emph{tycon signature} to it, which specifies the ``interface'' associated with the tycon structure.
+Here, line 2 states that the tycon is indexed by static values of \emph{kind} \liv{LblTyMap}.
+Kinds serve as the types of static values (discussed further in the next section).
+We assume that this kind has been defined previously  and specifies kind-specific syntax as mentioned when we described the record type \liv{Paper} above.
+
+The next two lines in the tycon signature specify two \emph{operator constructors} (or \emph{opcons}) associated with the tycon.
+The first is the introductory opcon.
+It is used when one writes a record literal, as in the body of \liv{exmpl_paper} above.
+The prefix \liv{ana} means that it can only be used in an analytic position, i.e.
+when an expected type is known.
+This is to ensure that the common set of introductory forms available in the language, like the record and string introductory forms shown being used in the examples above, can be unambiguously leveraged at many types.
+The semantics delegates to the tycon of the type that an expression of such a form is being analyzed against, as we will discuss in great detail below.
+The clause \liv{indexed by Lbl list} specifies that only introductory forms from which a list of statically valued labels can be derived are valid for introducing records.
+In this case, this means that only the record literal form is valid, but not the string literal form.
+
+The second opcon is written \liv{#}, and is used with the generalized projection form in the {\Verse} syntax.
+This is the syntax that is used in the example above to project the row \liv{conf} from \liv{paper}.
+The prefix \liv{syn} means that it synthesizes a type.
+The clause \liv{indexed by Lbl} indicates that a statically valued label, i.e.
+the name of the row being requested, must be provided to the opcon.
+The language delegates to the tycon of the type of the expression serving as the target of the projection.
+Again, we will discuss this in great detail below.
+
+\begin{figure}[t]\vspace{-5px}
+\begin{lstlisting}[numbers=left]
+tycon structure record ~ tcsig {
+  indexed by LblTyMap
+  ana intro indexed by Lbl list
+  syn # indexed by Lbl
+} = tcstruct {
+  translation = fn (tyidx : LblTyMap) -> ITy => 
+    fold tyidx `unit` (fn (lbl, ty) (rtr) => 
+      `trans(ty) * %rtr`)
+  ana intro = fn tyidx tmidx args => 
+    fold3 tyidx tmidx args `()` (
+      fn (lbl, ty) (rowlbl) (rowarg) (rtr) =>
+        let _ = lbleq lbl rowlbl
+        let rowtr = ana rowarg ty
+        `(%rowtr, %rtr)`)    
+  syn # = fn tyidx tmidx args => 
+    let arg0 = arity0 args
+    let (ty, pos) = lookup tyidx tmidx
+    (ty, generate_prj pos arg0)
+}
+\end{lstlisting}
+\caption{The tycon structure defining \liv{record} types.}
+\label{fig:record-conc}\vspace{-12px}
+\end{figure}
+
+The remainder of the listing gives the implementation of the tycon structure.
+The first component computes a \emph{type translation} for record types as a metafunction of the type index.
+Here, we choose to translate record types to nested binary product types by folding over the type index, with the base case being the nullary product type, \liv{unit}.
+Note that we are using quasiquotation syntax here to compute the type translation (another mode of use of TSLs).
+The form \liv{trans(ty)} refers \emph{abstractly} to the translation of another type.
+As we will see, the key to modular reasoning about tycon structures is \emph{translation independence}: that no other tycon structure can rely on the choice made here.
+The definition of records here could just as well use a different row ordering in its representation, or a different internal data structure entirely (e.g.
+a list instead of nested pairs).
+Here, the representation of the record type \liv{paper} will be \liv{trans(Title) * (trans(Conf) * unit)}.
+We could optimize the trailing unit away, but we leave it for simplicity and to again emphasize that this is merely an implementation detail.
+
+The remainder of the tycon structure determines the typechecking and translation logic governing the introductory and projection operators mentioned, and shown being used, above.
+We will return to the details of what is happening in subsequent sections.
+The main thing to notice for now is that types and translations are being statically computed.
+For example, one should be able to make out that nested pairs are being constructed in the intro opcon definition, and that we are projecting out the appropriate component based on the label provided to the projection opcon.
+Because projection is a synthetic opcon, it returns a pair of an (external) type and translation, whereas introduction returns only a translation.
+On line 13, it requests type analysis of an argument to the operation (i.e.
+a row value) against a corresponding type, showing how the typechecker is  exposed to the metalogic.
+
+To summarize: the semantics delegated control over 1) translation of the record type \liv{Paper}; 2) translation of the record literal in \liv{exmpl_paper} and 3) type synthesis and translation of the projection operator used to compute \liv{paper#conf} to the \liv{record} tycon structure, rather than specifying this logic \emph{a priori}.
+Similarly, though not shown here, the \liv{rstr} tycon was delegated control over 1) translation of the types \liv{Title} and \liv{Conf}; 2) translation of the string literals in the examples and 3) type synthesis and translation of the group projection operator.
+The language guaranteed that the choice of representation made by each tycon is known only to its opcons, so metatheoretic reasoning about whether, for example, \liv{rstr} correctly constrains its translation by the regular language provided as the type index is entirely local.
+
+%This suggests, however, that there is an opportunity to apply a principled approach to modularly reasoning about core language constructs that can be realized in this way.
+The Verse core language is organized like a compiler.
+
+% Programmers can choose from dialects supporting, e.g., a principled approach to distributed programming, or one that builds in support for statically reasoning about units of measure, one that supports both constructs may not be available.
 Using different dialects separately for different components of a program is untenable: components written in different dialects cannot always interface safely (i.e.
@@
 
-These problems with composition do not arise when a fragment is expressed as an isomorphic embedding (i.e.
-as a library) because modern \emph{module systems} can enforce abstraction barriers that ensure that the isomorphism needs only to be established in the ``closed world'' of the module.
-For example, a module for sets in ML can hold its representation type abstract, ensuring that any invariants necessary to maintain the isomorphism need only be maintained by the functions in the module (e.g.
-uniqueness, if using a list representation).
-These then continue to hold no matter which other modules are in use by a client \cite{harper1997programming}.% Other languages expose \emph{abstract data types} in other ways, also to localize reasoning \cite{liskov1974programming}.
-%Mechanisms that can help decrease the complexity of an embedding without violating abstraction barriers are thus valuable, and we will lead into our work in Sec.
-\ref{prior-work} by summarizing them.
-
-The alarming proliferation of dialects above suggests that mechanisms  that make it possible to define and  reason in a similarly modular, localized manner about  direct extensions to the semantics of a language are needed.
+%This puts the burden on the designers of general-purpose languages to maximize expressive power, so that dialect formation is less frequently necessary.
+%New language dialects are at best sources of ideas for the designers of languages that achieve critical mass.
+% %This suggests that researchers should put  more effort into integrating  ways to make that language as expressive as possible, so that new ideas like those found in dialects are  made available to programmers in a timely manner.
+% One way to do this is to attempt to constantly update a language with new but backwards compatible features via a community process. 
+% Arguably, Scala and Ocaml are archetypal examples of languages that successfully take this approach.
+Another is to strive toward a small number of highly general mechanisms, i.e.
+constructs that permit the isomorphic expression of constructs that other languages build in, or would need to build in.
+This paper introduces such a construct, called the \emph{tycon structure}.
+
+% The alarming proliferation of dialects above suggests that mechanisms  that make it possible to define and  reason in a similarly modular, localized manner about  direct extensions to the semantics of a language are needed.
 For example, once a language is extended with \emph{regular string types} as described in \cite{sanitation-psp14}, all terms having a regular string type like $\sty{\tcvar{rstr}}{\concrx{.+}}$ should continue to behave as non-empty strings no matter which other extensions are in use.
@@
 
-\paragraph{Contributions} In this paper, we take foundational steps toward this goal of modular language metatheory by  constructing a simple but surprisingly powerful  calculus, @$\lambda$ (the ``actively typed'' lambda calculus).
+% Researchers commonly invent minimal calculi or ``toy languages'' to describe and reason metatheoretically about interesting language constructs, leaving to future language designers the task of combining these to form a complete programming  language.
+% %Typed systems are often described as \emph{fragments}, each contributing separately to the concrete and abstract syntax and static and dynamic semantics of the language \cite{pfpl}. 
+% %It is common to organize such fragments around \emph{type constructors}, e.g.
+$\rightarrow$ for function types, describing each in a different chapter of his book.
+Languages are then identified by a set of type constructors, e.g.
+$\mathcal{L}\{{\rightharpoonup}\,{\forall}\,{\mu}\,{\keyw{1}}\,{\times}\,{+}\}$ is the language of partial (i.e.
+possibly non-terminating) function types, polymorphic types, recursive types, nullary and binary product types and binary sum types (its syntax is shown in Figure \ref{syntax-IL}, discussed below).
+% %Another common practice is to describe a fragment using a simple calculus having a ``catch-all'' constant and base type to stand notionally for all ``other'' terms and types.% (e.g.
+\cite{sanitation-psp14}).
+% 
+%This is not always a simple task.
+The usual metatheoretic reasoning techniques for programming languages  (e.g., rule induction) operate only on complete language specifications, so when two languages are combined, one must reestablish all metatheorems of interest anew, guided only informally by the metatheorems that were derived for these simpler systems.
+Indeed, this is not merely a rote exercise if we leave the language design space unconstrained, because not all language constructs make sense together.
+
+% \paragraph{Contributions} In this paper, we take foundational steps toward this goal of modular language metatheory by  constructing a simple but surprisingly powerful  calculus, $\lambda_\textsf{Verse}$ (the ``actively typed'' lambda calculus).
 %Despite its minimality, it can host a variety of practical semantic extensions like those described above, while maintaining strong metatheoretic guarantees and, crucially, providing modular reasoning properties.
@@
 \ref{overview}.
-Its semantics consist of an extensible \emph{external language} (EL) governed by a {typed translation semantics} targeting a simple and fixed \emph{internal language} (IL). 
-In particular, rather than building in a monolithic set of external type and operator constructors, the  semantics are indexed by a \emph{tycon context}.
+% Its semantics consist of an extensible \emph{external language} (EL) governed by a {typed translation semantics} targeting a simple and fixed \emph{internal language} (IL). 
+% In particular, rather than building in a monolithic set of external type and operator constructors, the  semantics are indexed by a \emph{tycon context}.
 Each tycon, e.g.
@@
 
-We call a tycon associated with opcons in this manner a \emph{modular tycon} because all \emph{translation invariants} maintained by the opcons associated with that tycon, e.g.
+% We call a tycon associated with opcons in this manner a \emph{modular tycon} because all \emph{translation invariants} maintained by the opcons associated with that tycon, e.g.
 the regular string invariant just mentioned, are necessarily maintained in any further extended tycon context, i.e.
@@
 
-% We will begin by giving an overview of  @$\lambda$ in Sec.
+% We will begin by giving an overview of  $\lambda_\textsf{Verse}$ in Sec.
 \ref{atlam}, introducing our two main examples, one defining labeled product types (i.e.
@@
 
-\section{Overview of @$\lambda$}\label{atlam}\label{overview}
+%\vspace{5px}
+\section{Overview of $\lambda_\textsf{Verse}$}\label{atlam}\label{overview}
+Let us now make the intuitions in the previous section completely precise, by specifying a core calculus, $\lambda_\textsf{Verse}$.
+We cover the most important rules in the paper, but a significantly more detailed treatment is provided in the accompanying supplement.
 \begin{figure}[t]
@@
 \hspace{-5px}\input{syntax-table-IL}
-\caption{Syntax of {$\mathcal{L}\{{\rightharpoonup}\,{\forall}\,{\mu}\,{\iunit}\,{\times}\,{+}\}$}, our internal language (IL).
-Metavariable $x$ ranges over term variables and $\alpha$ and $t$ (distinguished only for stylistic reasons) over type variables.}
+\caption{Syntax of our internal language (IL). Metavariable $x$ ranges over term variables and $\alpha$ (or $t$) over type variables.}
 \label{syntax-IL}
@@
 We use {$\mathcal{L}\{{\rightharpoonup}\,{\forall}\,{\mu}\,{\iunit}\,{\times}\,{+}\}$}, Figure \ref{syntax-IL}, as representative of a typical intermediate language for a typed language.
-%Note that the IL has a fixed semantics; changes to the IL induce different dialects of @$\lambda$.
+%Note that the IL has a fixed semantics; changes to the IL induce different dialects of $\lambda_\textsf{Verse}$.
 We assume an internal statics specified by judgements for type assignment {$\iDelta~\iGamma \vdash \iota : \tau\moutput$}, type formation {$\iDelta \vdash \tau$} and typing context formation { $\iDelta \vdash \iGamma$}, and an
@@
 
-\paragraph{External Language} Programs ``execute'' as internal terms, but programmers interface with @$\lambda$ by writing \emph{external terms}, $e$.
+\paragraph{External Language} Programs ``execute'' as internal terms, but programmers interface with $\lambda_\textsf{Verse}$ by writing \emph{external terms}, $e$.
 The abstract syntax of external terms is shown in Figure \ref{syntax-EL} and we introduce various concrete desugarings as we go on.
@@
 record types, by translation to the IL.
-%In @$\lambda$, the EL builds in only function types. All others arise from the \emph{tycon context}, $\Phi$.
-
-%The reason is that in the Harper-Stone elaboration semantics, external and internal terms were governed by a common type system, but in @$\lambda$
+%In $\lambda_\textsf{Verse}$, the EL builds in only function types. All others arise from the \emph{tycon context}, $\Phi$.
+
+%The reason is that in the Harper-Stone elaboration semantics, external and internal terms were governed by a common type system, but in $\lambda_\textsf{Verse}$
 
@@
 \end{figure}
-As suggested above, the main novelty of @$\lambda$ is that the types and term and type translations do not arise from a fixed specification.
+As suggested above, the main novelty of $\lambda_\textsf{Verse}$ is that the types and term and type translations do not arise from a fixed specification.
 Rather, they are \emph{statically computed} by tycon definitions using a \emph{static language} (SL).
@@
 The kinding judgement takes the form $\sofkX{\st}{\kappa\moutput}$, where $\kDelta$ and $\kGamma$ are analagous to $\iDelta$ and $\iGamma$ and analagous judgements $\kDelta \vdash \kappa$ and $\kDelta \vdash \kGamma$ are  defined.
-% All such contexts in @$\lambda$ are identified up to exchange and contraction and obey weakening \cite{pfpl}.
+% All such contexts in $\lambda_\textsf{Verse}$ are identified up to exchange and contraction and obey weakening \cite{pfpl}.
 The natural number $n$ is simply a bound used to prevent ``out of bounds'' references to arguments. 
@@
     \sofkX{\sty{\keyw{other}[m; \kappa]}{\st}}{\kty}
-}\vspace{-10px}
+}%\vspace{-10px}
 \end{mathpar}
@@
 \end{figure*}
-\section{Types}\label{types}
+\paragraph{Types}\label{types}
 
@@
 
-\noindent
 Types are static values of kind $\kty$, i.e.
@@
     \vdash \Phi, \tcdef{\tc}{\tcsig{\ktyidx}{\chi}}{\tcstruct{\stx{schema}}{\omega}}
-}\vspace{-10px}
+}%\vspace{-10px}
 \end{mathpar}
@@
     \sofkX{\ssyn{n'}}{\kty\times\kitm}
-}\vspace{-10px}
+}
 \end{mathpar}
@@
 \begin{figure}
-\small\fbox{$\trvalidate{\argEnv}{c}{\qitm}{\st}{\iota\moutput}$}\vspace{-5px}
+\small\fbox{$\trvalidate{\argEnv}{c}{\qitm}{\st}{\iota\moutput}$}
 \begin{mathpar}\small
@@
 \trvalidate{\es;\Upsilon;\Phi}{c}{\qitm}{\st}{[\delta][\gamma]\iota_\text{abs}}
-}\end{mathpar}\vspace{-9px}
+}\end{mathpar}
 \caption{Translation Validation}
@@
 
-Whereas introductory operations were analytic, targeted operations are synthetic in @$\lambda$.
+Whereas introductory operations were analytic, targeted operations are synthetic in $\lambda_\textsf{Verse}$.
 The type and translation are determined by the tycon of the type synthesized by the target argument.
@@
 \section{Metatheory}\label{metatheory}
-\noindent We will now state the key metatheoretic properties of @$\lambda$. The proofs are in the supplement.
+\noindent We will now state the key metatheoretic properties of $\lambda_\textsf{Verse}$. The proofs (with a few straightforward details left to be fleshed out) are in the supplement.
 
@@
 
-\section{Related Work and Discussion}\label{prior-work}
+\section{Related Work and Discussion}\label{prior-work}\label{sec:discussion}
 \noindent We are not the first to use a semantics distinguishing the EL from a smaller IL  for a language specification.
@@
 
-In contrast, in @$\lambda$, fragment providers need not provide the semantics with mechanized specfications or proofs to benefit from rigorous modular reasoning principles.
+In contrast, in $\lambda_\textsf{Verse}$, fragment providers need not provide the semantics with mechanized specfications or proofs to benefit from rigorous modular reasoning principles.
 Instead, under a fixed tycon context, the calculus can be reasoned about like a very small type-directed compiler \cite{tarditi+:til-OLD,conf/pldi/Chlipala07,Dave:2003:CVB:966221.966235}.
@@
 To eliminate even these localized failure modes, we plan to introduce \emph{optional} specification and proof mechanization into the SL (by basing it on a dependently typed language like Coq, rather than ML).
-Because types are values in the SL, the SL can be seen as building on the concept of \emph{type-level computation}\todo{find appropriate citation}.
-%We also hope to mechanize the metatheory of @$\lambda$ itself in future work.%Every combination of fragments is a new dialect, and must be reasoned about monolithically.
+Because types are values in the SL, the SL can be seen as building on the concept of \emph{type-level computation}.
+%We also hope to mechanize the metatheory of $\lambda_\textsf{Verse}$ itself in future work.%Every combination of fragments is a new dialect, and must be reasoned about monolithically.
 %Efforts to reason modularly areeed, several problems can come up when fragments are combined, so .
@@
 In contrast, modules and ADTs can only specify a fixed number of operations, and each must have function type.
-Note also that these are not competing mechanisms: we did not specify quantification over external types here for simplicity, but we conjecture, based on the finding that polymorphism is conservative over simple types, \cite{breazu1990polymorphism}, that this is  complementary and thus @$\lambda$ could serve as the core of a language with an ML-style module system and polymorphism.
+Note also that these are not competing mechanisms: we did not specify quantification over external types here for simplicity, but we conjecture, based on the finding that polymorphism is conservative over simple types, \cite{breazu1990polymorphism}, that this is  complementary and thus $\lambda_\textsf{Verse}$ could serve as the core of a language with an ML-style module system and polymorphism.
 Another related direction is to explore \emph{tycon functors}, which would abstract over tycons with the same signature to support modularly tunable cost semantics.
@@
 \section{Conclusion}
-\noindent We have specified a simple but surprisingly powerful typed translation semantics, @$\lambda$, where new type constructors can be introduced ``from within''.
+\noindent We have specified a simple but surprisingly powerful language, {\Verse}, specified at its core by a typed translation semantics, $\lambda_\textsf{Verse}$, where new type constructors can be introduced ``from within''.
 The corresponding operator constructors determine types and translations ``actively'', i.e.
@@
 a \opname{len} opcon on regular strings could not construct a natural number).
-We conjecture that these are not fundamental limitations and expect @$\lambda$ to serve as a  foundation for future efforts that increase  expressiveness while maintaining the strong modularity guarantees established  here.
+We conjecture that these are not fundamental limitations and expect $\lambda_\textsf{Verse}$ to serve as a  foundation for future efforts that increase  expressiveness while maintaining the strong modularity guarantees established  here.
 %(tycons could define new contexts which are threaded opaquely through ``outside'' operations).
@@
 
-\newpage
 % We recommend abbrvnat bibliography style.

### 2015-05-28 atlam-icfp15/att-pldi15-supplement.tex
ICFP 15

--- 2014-11-25 att-pldi15/att-pldi15-supplement.tex
@@
 
-\title{Modularly Composing Typed Language Fragments (Supplemental Material)}
+\title{Modularly Metaprogrammable Type Structure (Supplemental Material)}
 %\subtitle{Supplemental Material}
@@
 \abstract{
-    This document provides the full technical development described in the paper ``Modularly Composing Typed Language Fragments'' submitted to PLDI 2015. 
+    This document provides more complete technical development described in the paper ``Modularly Metaprogrammable Type Structure'' submitted to ICFP 2015. 
 }
@@
 \subsection{Internal Statics}
-We assume an internal language where the statics are specified in the standard way by judgements for  type formation {$\iDelta \vdash \tau$}, shown in Figure \ref{fig:internal-type-formation}, typing context formation { $\iDelta \vdash \iGamma$}, shown in Figure \ref{fig:internal-typing-context-formation}, and type assignment {$\iDelta~\iGamma \vdash \iota : \tau\moutput$}, shown in Figure \ref{fig:internal-type-asignment}.
-\todo{do these and cite relevant rules in remainder}
+We assume an internal language where the statics are specified in the standard way by judgements for  type formation {$\iDelta \vdash \tau$}  typing context formation { $\iDelta \vdash \iGamma$}, and type assignment {$\iDelta~\iGamma \vdash \iota : \tau\moutput$}.
 %The typing and type formation contexts obey standard structural properties (i.e. weakening, exchange and contraction).
@@
 \subsubsection{Backprojection}
-To avoid having to give a large number of rules that do nothing interesting, we define backprojection on static terms definitionally.\todo{eval of index should be empty Phi}
+To avoid having to give a large number of rules that do nothing interesting, we define backprojection on static terms definitionally.%\todo{eval of index should be empty Phi}
 
@@
 
-This section has, obviously, been less detailed than the remainder of the supplement, essentially because the lemmas above involve a rather large number of cases where nothing of note is happening, and we simply ran out of time to inductively specify the full details of backprojection and prove all of them.
+This section needs a bit more detail, essentially because the lemmas above involve a rather large number of cases where nothing of note is happening, and we simply ran out of time to inductively specify the full details of backprojection and prove all of them.
 Backprojection is a rather simple conceptual mapping, so we believe that the level of detail above is sufficiently convincing, and we plan on writing out the full details (which we anticipate will be another 20 or so pages) well before the notification date.
@@
 \section{Discussion}
-\subsection{Record Types}
-In the paper, we presented labeled products rather than the more conventional record types, both to emphasize that different variants have different trade-offs, and because it made our examples simpler.
-
-One strategy for supporting records themselves would be to specify the intro operator such that records that do not have sorted indices are uninhabited, then provide a helper static function that generates only sorted indices and require that all type construction pass through that.
+% \subsection{Record Types}
+% In the paper, we presented labeled products rather than the more conventional record types, both to emphasize that different variants have different trade-offs, and because it made our examples simpler.
+
+% One strategy for supporting records themselves would be to specify the intro operator such that records that do not have sorted indices are uninhabited, then provide a helper static function that generates only sorted indices and require that all type construction pass through that.
 This notion of a ``type constructor constructor'' could perhaps be integrated into the calculus.
 
-An alternative strategy may have been to use an abstract kind that ensured that such type indices could not have been constructed, but to be compatible with our equality kind restriction, this would require support for abstract equality kinds, analagous to abstract equality types in SML.
+% An alternative strategy may have been to use an abstract kind that ensured that such type indices could not have been constructed, but to be compatible with our equality kind restriction, this would require support for abstract equality kinds, analagous to abstract equality types in SML.
 We chose not to formalize these for simplicity, and to demonstrate the general technique above.

### 2015-05-28 thesis-proposal/problem.tex
Final draft

@@
 Instead, its users are able to modularly  express a wide variety of useful constructs in libraries, using the comparatively small set of primitives that the language builds in.
-Functional programming languages like Standard ML (SML), OCaml and Haskell descend from mathematically elegant typed lambda calculi, diverging mainly in that they consider various pragmatic issues alongside  fundamental metatheoretic issues in their design.
-For example, they build in record types, generalizing the nullary and binary product types more common in minimal calculi, because labeled components are often cognitively useful.
-Similarly, they build in derived concrete syntax for commonly used constructs like lists.% concision and readability.
+Functional programming languages like Standard ML (SML), OCaml and Haskell descend from mathematically elegant typed lambda calculi, diverging mainly in that they consider a number of pragmatic issues alongside  fundamental metatheoretic issues in their design.
+For example, they build in record types, generalizing the nullary and binary product types more common in minimal calculi, because explicitly labeled components can be cognitively useful.
+Similarly, they build in derived concrete syntax for common constructs like lists.% concision and readability.
 %Minimal typed lambda calculi play a central role in the study of programming languages, and serve as direct inspiration for  functional programming languages like ML.
@@
 
-Ideally, only a limited number of ``embellishments'' like these would be needed to turn a minimal lambda calculus into a practical ``full-scale'' functional programming language.
-The community around the resulting language would then be able to direct its efforts almost exclusively toward the development of useful libraries, optimizing compilers and other tools.
-Unfortunately, a stable language design that fully achieves this ideal has yet to emerge, as evidenced by the diverse array of \emph{dialects} -- languages derived from a common base language that introduce different embelishments -- that continue to proliferate around all major contemporary languages.
+The hope is that a limited number of ``embellishments'' like these will suffice to turn a minimal lambda calculus into a broadly practical programming language.
+%The community around this language should, ideally, then be able to direct its efforts almost exclusively toward the development of useful libraries, rather than further modifications to the language itself.
+Unfortunately, a stable language design that actually achieves this ideal has yet to emerge, as evidenced by the diverse array of \emph{dialects} that continue to proliferate around all major contemporary languages.
 % For example, let us consider the many dialects of ML.
@@
 %Tools for  constructing new ``domain-specific'' dialects also continue to proliferate. % The situation is similar across all major language lineages. 
-In fact, tools that assist with the construction of so-called  ``domain-specific'' language dialects seem only to be increasingly popular. 
-{This calls for an investigation}: why are programmers and researchers so often unable to satisfyingly express the constructs that they need in libraries, as modes of use of the ``general-purpose'' primitives already available in major functional languages?
+In fact, tools that assist with the construction of so-called  ``domain-specific'' language dialects are only becoming increasingly popular. 
+{This calls for an investigation}: why is it that programmers and researchers are so often unable to satisfyingly express the constructs that they need in libraries, as modes of use of the ``general-purpose'' primitives already available in major languages today?
 %Let us investigate why this is.
@@
 
-Perhaps the most common reason may simply be that the \emph{syntactic cost} of expressing a construct of interest using contemporary general-purpose primitives is not always ideal.
-To decrease syntactic cost, library providers are often tempted to construct \emph{derived syntactic dialects}, i.e.
+Perhaps the most common reason for this proliferation of dialects may simply be that the \emph{syntactic cost} of expressing a construct of interest using contemporary general-purpose primitives is not always ideal.
+To decrease syntactic cost, library providers construct \emph{derived syntactic dialects}, i.e.
 dialects that are specified by context-free elaboration to the existing language.
-For example, Ur/Web is a  derived syntactic dialect of Ur (a language that descends from ML) that builds in derived syntax (colloquially, ``syntactic sugar'') for SQL queries, HTML elements and other datatypes used for web programming \cite{conf/popl/Chlipala15}.
+For example, Ur/Web is a  derived syntactic dialect of Ur (a language that itself descends from ML) that builds in derived syntax (colloquially, ``syntactic sugar'') for SQL queries, HTML elements and other datatypes used for web programming \cite{conf/popl/Chlipala15}.
 %Syntactic cost is often assessed qualitatively \cite{green1996usability}, though quantitative metrics can be defined.
-Many other types of data similarly benefits from the availability of specialized derived syntax.
+Many other types of data can similarly benefit from the availability of specialized derived syntax.
 For example, we will consider regular expression patterns expressed using abstract data types (as a mode of use of the ML module system) in Sec.
 \ref{sec:syntax}.
-Though we will be able to express  precisely the semantics that we seek,   the usual approach of using dynamic string parsing to introduce a pattern will leave something to be desired syntactically.
+We will be able to express precisely the semantics that we seek, but the usual approaches, e.g.
+using dynamic string parsing, for approximating the concrete syntax of patterns will leave something to be desired.
 % An extension of Moscow ML provided syntax for monads encoded using the module system, inspired by the  syntax built in to Haskell.\todo{asked for reference, if not found will say that "Harper reports..."} %In fact, nearly all major programming languages primitively build in derived syntax for constructs that can otherwise be expressed in the standard library, e.g.
 derived list syntax in SML \cite{harper1997programming,mthm97-for-dart}.
-Tools like Camlp4 \cite{ocaml-manual} and Sugar* \cite{erdweg2011sugarj,erdweg2013framework} are designed to lower the engineering costs of constructing derived syntactic dialects in situations like these, contributing to their continued proliferation.
+Tools like Camlp4 \cite{ocaml-manual} and Sugar* \cite{erdweg2011sugarj,erdweg2013framework} have lowered the engineering costs of constructing derived syntactic dialects in situations like these, contributing to their ongoing proliferation.
 %(we will consider them further in Sec.
@@
 
-More advanced dialects define new static type structure, going beyond what is possible by context-free elaboration to the existing language.
-Record types themselves are perhaps  the simplest example -- they cannot be expressed by context-free elaboration to a language with only nullary and binary products.
-Various dialects of ML and Haskell have explored  ``record-like'' primitives that go still further, supporting functional update and extension operators, width and depth coercions (sometimes implicit)%\cite{Cardelli:1984:SMI:1096.1098}
-, methods, prototypic dispatch and other ``semantic embellishments'' that cannot be expressed by context-free elaboration to a language with only standard record types (we will detail one  example in Sec.
+More advanced dialects introduce new type structure, going beyond what is possible by context-free elaboration to the existing language.
+Record types are themselves a simple example -- they cannot be expressed by context-free elaboration to a language with only nullary and binary product types.
+Various functional languages have explored ``record-like'' primitives that go still further, supporting functional update and extension operators, width and depth coercions (sometimes implicit)%\cite{Cardelli:1984:SMI:1096.1098}
+, methods, prototypic dispatch and other ``semantic embellishments'' that cannot be expressed by context-free elaboration to a language with only standard record types (we will detail an  example in Sec.
 \ref{sec:metamodules}).
-The OCaml dialect of ML primitively builds in ``polymorphic variants'', ``open datatypes'' and  logic for typechecking operations that use format strings like $\mathtt{sprintf}$ \cite{ocaml-manual}.
+OCaml primitively builds in the type structure of ``polymorphic variants'', ``open datatypes'' and  operations that use format strings like $\mathtt{sprintf}$ \cite{ocaml-manual}.
 ReactiveML builds in primitives for functional reactive programming \cite{mandel2005reactiveml}.
 ML5 builds in high-level primitives for distributed programming based on a modal lambda calculus \cite{Murphy:2007:TDP:1793574.1793585}.
-Manticore \cite{conf/popl/FluetRRSX07} and Alice ML  \cite{AliceLookingGlass} build in parallel programming primitives with a more elaborate type structure than is found in their respective core calculi.
+Manticore \cite{conf/popl/FluetRRSX07} and AliceML  \cite{AliceLookingGlass} build in parallel programming primitives with a more elaborate type structure than is found in their respective core calculi.
 MLj builds in the type structure of the Java object system (motivated by a desire to interface safely and naturally with Java libraries) \cite{Benton:1999:IWW:317636.317791}.
@@
 Second, primitives introduced earlier in a language's lifespan end up monopolizing finite ``syntactic resources'', forcing subsequent primitives to use ever more esoteric forms.
-A third problem is that primitives cannot be removed or changed without breaking backwards compatibility.
+And third, primitives that prove to be flawed in some way cannot be removed or changed without breaking backwards compatibility.
 %rimitives that are only situationally useful, or that trigger aesthetic disagreements between different factions of the community, are still quite often left languishing in impractical ``toy'' dialects.
@@
 %This approach is antithetical to the ideal of a truly \emph{general-purpose language} described at the beginning of this section.
-
+\newpage
 \begin{quote}\textit{The recent development of programming languages suggests that the simul\-taneous achievement of simplicity 
@@
 For example, a recent revision of Ocaml added support for  ``generalized algebraic data types'' (GADTs), based on research on guarded recursive datatype constructors \cite{XiCheChe03}.
-Using GADTs, Ocaml was able to move some of the \emph{ad hoc} machinery for typechecking operations that use format strings, like \texttt{sprintf}, out of the language and into a library (note however that syntactic machinery remains built in).
+Using GADTs, Ocaml was able to move some of the \emph{ad hoc} machinery for typechecking operations that use format strings, like \texttt{sprintf}, out of the language and into a library (note however that syntactic machinery remains primitively built in).
 
@@
 \section{Proposed Contributions}
-Our broad aim in the work being proposed is to introduce a new functional programming language called Verse\footnote{We distinguish Verse from Wyvern, which is the language referred to in prior publications about some of the work being proposed here, because Wyvern is a group effort evolving independently in some important ways.} that takes more radical steps down the second path just described by giving library providers the ability to directly express derived concrete syntax and type structure of a variety of designs in a safe and modularly composable manner.
-%This eliminates the needs to build in fewer \emph{ad hoc} constructs and dialects are less frequently necessary.
+Our broad aim in the work being proposed is to introduce a new functional programming language called Verse\footnote{We distinguish Verse from Wyvern, which is the language referred to in prior publications about some of the work being proposed here, because Wyvern is a group effort evolving independently in some important ways.} that takes more radical steps down the second path just described by giving library providers the ability to directly express new derived concrete syntax and type structure of a variety of designs in a safe and modularly composable manner atop a minimal type-theoretic core.
+The result is that Verse is smaller than comparable languages like ML and Scala, and it renders dialect formation less frequently necessary.%This eliminates the needs to build in fewer \emph{ad hoc} constructs and dialects are less frequently necessary.
 %thereby reducing the need for language dialects and revisions.
 
-Verse features a module system taken directly from SML.
+%Verse features a module system taken directly from SML.
 Unlike SML, the Verse core language is split into a \emph{typed external language} (EL) specified by {type-directed translation} to a minimal \emph{typed internal language} (IL).
-We formally introduce the semantics of the Verse core language in Sec.
-\ref{sec:verse}.
-Our novel contributions will be two primitive constructs that eliminate the need for many others:
-\begin{itemize}
+We formally introduce the semantics of Verse in Sec. \ref{sec:verse}. Our novel contributions will be two primitives that eliminate the need for many others:
+\begin{enumerate}
 \item \textbf{Typed syntax macros} (TSMs), introduced in Sec.
 \ref{sec:syntax}, eliminate the need to primitively build in derived concrete syntax specific to library constructs, e.g.
-list syntax as in SML or XML syntax as in Scala and Ur/Web, by giving library providers static control over the parsing and elaboration of delimited segments of concrete syntax.
+list syntax as in SML or XML syntax as in Scala and Ur/Web, by giving library providers static control (at a specified type or parameterized family of types) over the parsing and elaboration of delimited segments of concrete syntax.
 %We show how TSMs can be invoked explicitly by clients, or, implicitly based on local type inference.
 \item \textbf{Metamodules}, introduced in Sec.
-\ref{sec:metamodules}, eliminate the need to primitively build in the type structure of constructs like records (and variants thereof),  labeled sums and other more esoteric constructs that we will consider by giving library providers static hooks directly into Verse's type-directed translation semantics.
-For example, a library provider can implement the type structure of records by defining a metamodule that:
-\begin{enumerate}
-\item introduces a type constructor, \lstinline{record}, parameterized by finite mappings from labels to types, and defines its semantics by translation to unary and binary products (which are built in to the internal language); and
-\item introduces operators used to work with records, minimally record introduction and elimination (but perhaps also various functional update operators), and directly defines the logic governing their typechecking and translation to the IL (which builds in only nullary and binary products).
+\ref{sec:metamodules}, eliminate the need to primitively build in the type structure of constructs like records (and variants thereof),  labeled sums and other more esoteric constructs that we will introduce later by giving library providers programmatic hooks directly into the semantics.
+%For example, a library provider can implement the type structure of records with a metamodule that:
+%\begin{enumerate}
+%\item introduces a type constructor, \lstinline{record}, parameterized by finite mappings from labels to types, and defines, programmatically, a translation to unary and binary product types (which are built in to the internal language); and
+%\item introduces operators used to work with records, minimally record introduction and elimination (but perhaps also various functional update operators), and directly implements the logic governing their typechecking and translation to the IL (which builds in only nullary and binary products).
+%\end{enumerate}
+We will see direct analogies between ML-style modules (which Verse also supports) and metamodules in Sec. \ref{sec:metamodules}.
 \end{enumerate} 
-We will see direct analogies between modules and metamodules in Sec. \ref{sec:metamodules}.
-\end{itemize} 
-Both TSMs and metamodules involve \emph{static metaprogramming}, i.e.
-their static semantics will involve the evaluation of user-defined functions that manipulate static representations of types and terms.% static code generation.
+Both TSMs and metamodules are novel forms of \emph{static code generation}, i.e.
+the relevant rules in the static semantics call for the evaluation of \emph{static functions} that generate static representations of expressions and types.
+These are written in a \emph{static language} (SL).
+% static code generation.
 %We will  also introduce a simple variant of each of these primitives that leverages Verse's support for local type inference to further reduce syntactic cost in certain common situations.
 
-The key challenge in the design of these primitives will come in ensuring that they are metatheoretically well-behaved, given that they aim to decentralize control over aspects of a language's concrete syntax and type structure that, in other contemporary languages, are under the exclusive control of the language designer.
+The key challenge in the design of these primitives will come in ensuring that they are metatheoretically well-behaved, given that they aim to decentralize control over aspects of the concrete syntax and type structure that, in other contemporary languages, are under the exclusive control of the language designer.
 If we are not careful, many of the problems  that arise when combining language dialects, discussed earlier, could simply shift into the semantics of these primitives.\footnote{This is why languages  like Verse are often called \emph{extensible languages}, though this is somewhat of a misnomer.
 The chief characteristic of an extensible language is that it \emph{doesn't} need to be extended in situations where other languages would need to be extended.
-We will avoid this somewhat confusing terminology.} Our main technical contributions will come in showing how to address these problems in a principled manner.
+We will avoid this somewhat confusing terminology.} Our main technical contributions will be in showing how to address these problems in a principled manner.
 In particular, syntactic conflicts will be impossible by construction and the semantics will validate code statically generated by TSMs and metamodules to maintain a strong \emph{hygienic type discipline} and, most uniquely, powerful \emph{modular reasoning principles}.
@@
 We will tacitly assume that they are being avoided extrinsically, e.g.
-by using a URI-based naming scheme as in the Java ecosystem.}
+by using a URI-based naming scheme as in the Java ecosystem.} We will make these notions completely precise as we continue.
 
@@
 \begin{quote}
-A functional programming language can give library providers the ability to meta\-pro\-gram\-matic\-ally express new derived syntax and external type structure atop a minimal typed internal language while maintaining a hygienic type discipline and modular reasoning principles.
+A functional programming language can give library providers the ability to %meta\-pro\-gram\-matic\-ally 
+implement new derived concrete syntax and type structure atop a minimal type-theoretic core while maintaining a hygienic type discipline and modular reasoning principles.
 %These  primitives are  expressive enough to subsume the need for a variety of primitives that are, or would need to be, built in to comparable contemporary languages.
@@
 \subsection{Disclaimers}
-Before we continue, let us explicitly acknowledge that completely eliminating the need for dialects would indeed be asking for too much: certain design decisions are fundamentally incompatible with others or require coordination across a language design.
+Before we continue, it may be useful to explicitly acknowledge that completely eliminating the need for dialects would indeed be asking for too much: certain design decisions are fundamentally incompatible with others or require coordination across a language design.
 We aim only to decrease the need for dialects.% out a larger design space within a single language, Verse.%a subset of constructs that can be specified by a semantics of a certain ``shape'' specified by Verse (we will make this more specific later).
@@
 We  expect that in practice, Verse will come with a standard library defining a carefully curated collection of standard constructs, as well as guidelines for advanced users regarding when it would be sensible to use the mechanisms we introduce (following the example of languages that support operator overloading or typeclasses, which also have the potential for ``abuse'').
-%The vast majority of programmers should not use the primitives that we introduce directly.
-
-Finally, we will not be interested here in languages that feature full-spectrum dependent types, which blur the phase separation between ``compile-time'' and ``run-time'', though we conjecture that the primitives we describe could be added to languages like Gallina (the ``external language'' of the Coq proof assistant) with some suitable modifications.
-Verse is designed for use in situations where ML, Haskell or Scala would be suitable today.
+For most programmers, using Verse should not be substantially different from using a comparable language or one of its dialects.%The vast majority of programmers should not use the primitives that we introduce directly.
+
+Finally, Verse intentionally does not support full-spectrum dependent types because they blur the phase separation between ``compile-time'' and ``run-time''.
+Verse is designed for use in situations where SML, Ocaml, Haskell or Scala would be used today.
+That said, we conjecture that the primitives we describe could be added to languages like Gallina (the ``external language'' of the Coq proof assistant) with some modifications.
 %Our interest is in making sure that the standard library is no more privileged than any other library.
@@
 
-\section{Overview of Judgements}\label{sec:verse}
-To situate ourselves within a formal framework, let us begin with a brief overview of how Verse is organized and specified. %Verse consists of a \emph{module language} atop a \emph{core language}.
-
-\subsection{Module Language}
-The Verse module language is taken directly from Standard ML, with which we assume a working familiarity for the purposes of this proposal \cite{harper1997programming,MacQueen:1984:MSM:800055.802036} (a related module language, e.g.
-Ocaml's, would work just as well for our purposes).
-We will give examples of its use in Sections \ref{sec:syntax} and \ref{sec:metamodules}, but because it has been thoroughly studied in the literature, we will defer to prior work both here and in the dissertation for the formal details.% Instead, we will consider it abstractly where it interacts with the core language.%\todo{reminder that we are not going to re-specify the module system}
+\section{Verse}\label{sec:verse}
+To situate ourselves within a formal framework, let us begin with a brief overview of how Verse is organized and specified. Verse consists of a \emph{module language} atop a \emph{core language}.
 
 \subsection{Core Language}
-The underlying core language -- the language of types and expressions -- will be the focus of our formal efforts.
-The Verse core language is organized much like the first stage of a type-directed compiler (e.g.
-the TIL compiler for Standard ML \cite{tarditi+:til-OLD}), consisting of a user-facing \emph{typed external language} (EL) specified by type-directed translation to a minimal \emph{typed internal language} (IL).
+The Verse core language -- the language of types and expressions -- will be the focus of our formal efforts.
+The key novelty is that the core language is split into a user-facing \emph{typed external language} (EL) specified by type-directed translation to a minimal \emph{typed internal language} (IL).
+Notionally, one might compare the core language to the first stage of a type-directed compiler (e.g.
+the TIL compiler for Standard ML \cite{tarditi+:til-OLD}) shifted ``one level up'' into the language itself.
+A third sublanguage, the \emph{typed static language} (SL), is involved in giving library providers programmatic control over aspects of this type-directed translation from the EL to the IL.
+% The SL is notionally analagous to the language that the first stage itself is written in.
+
+We summarize the main judgements relevant to the IL, EL and SL below. Readers who prefer to see examples first can safely skim these sections for now.
 
 \subsubsection{Internal Language}
-Let us begin with the internal language.
-For our purposes, we keep things simple by using the strictly evaluated polymorphic lambda calculus with nullary and binary product and sum types and recursive types.
-We assume here that the reader has familiarity with these (we follow \emph{PFPL} \cite{pfpl} directly).
-To briefly review, the main judgements in the static semantics of the IL take the following standard form (omitting contexts for now):
+Programs ultimately evaluate as \emph{internal expressions}, $\iota$.
+Internal expressions are classified by \emph{internal types}, $\tau$, so the internal language forms a standard typed lambda calculus.
+For our purposes, it suffices to use the strictly evaluated polymorphic lambda calculus with nullary and binary product and sum types and recursive types as our IL.
+We assume in this proposal that the reader is familiar with this language (we follow \emph{PFPL} \cite{pfpl} directly).
+The main judgements in the static semantics of the IL take the following familiar form (omitting contexts for simplicity throughout this section):
 \\[1ex]
@@
 \noindent
-The dynamic semantics can be specified also in the standard manner as a transition system with  judgements of the following form:
+The dynamic semantics are specified also in the standard manner as a transition system with  judgements of the following form:
 \\[1ex]
@@
 
-Features like state, exceptions, minimal concurrency primitives, scalars, arrays and others characteristic of a first-stage compiler intermediate language would also be included in the IL in practice, and in some of these cases, this would affect the shape of the internal semantics.
-However, our design is largely insensitive to the details of the internal language -- the only strict requirements are that the IL be type safe and support parametric type abstraction -- so we will stick to this simpler core here.
+Note that features like state, exceptions, minimal concurrency primitives, scalars, arrays and other primitives characteristic of a first-stage compiler intermediate language would also be included in the IL in practice, and for several of these, this would affect the shape of the internal semantics.
+However, our design is largely insensitive to such details -- the only strict requirements are that the IL be type safe and support parametric type abstraction.
+We wish to give a minimal account of our novel contributions, so we will stick to this simpler, well-understood IL for the purposes of this work.
 %Features like record types and labeled sum types, on the other hand, are not built in.
@@
 \subsubsection{External Language}
-Programs evaluate as internal expressions but they are written as external expressions.
-For concision, we will use the terms ``type'' and ``expression'' without qualification to refer to  \emph{external types}, $\sigma$,  and \emph{external expressions}, $e$, respectively.
-Note that external types are distinct from internal types.
+The external language supports a richer syntax and type structure atop the IL.
+Throughout this work, the words ``expression'' and ``type'' used without qualification refer to external expressions and types, respectively.
 The main judgements in the specification of the EL take the following form  (again omitting various contexts for now):
@@
 $\\[1ex]
-Note that the expression typing judgements are \emph{bidirectional}, i.e.
+Notice that the expression typing judgements are \emph{bidirectional}, i.e.
 we make a judgemental distinction between \emph{type synthesis} (the type is an ``output'') and \emph{type analysis} (the type is an ``input'') \cite{Pierce:2000:LTI:345099.345100}.
-This is to  allow us to explicitly specify how Verse's \emph{local type inference} works, and in particular, how it interacts with the primitives that we aim to introduce.
-Like Scala, we intentionally do not attempt to support global type inference.
+This will  allow us to explicitly specify how Verse's \emph{local type inference} works, and in particular, how it interacts with the primitives that we aim to introduce.
+Like Scala, we intentionally do not seek to support global type inference.
 %, primarily because it is not technically compatible with the primitives we will introduce (but also because we believe that compilers for languages that support only local type inference generate clearer error messages).
+
+\subsubsection{Static Language}
+Finally, the static language plays an essential role in the semantics of the novel primitives that we will introduce into the EL.
+The SL is another typed lambda calculus consisting of \emph{static expressions}, $\sigma$, classified by \emph{sorts}\footnote{We use this word to avoid confusion with the common (though redundant) phrase ``static type''.}, $\uppi$, according to the following judgements (the ``static statics''):
+\\[1ex]
+$
+\begin{array}{ll}
+\textbf{Judgement Form} & \textbf{Pronunciation}\\
+\vdash \sigma :: \uppi & \text{Static expression $\sigma$ has sort $\uppi$.}\\
+\vdash \uppi~\mathtt{sort} & \text{Sort $\uppi$ is valid.}
+\end{array}
+$\\
+
+\noindent
+The ``static dynamics'' are specified as a transition system with judgements of the following basic form (though again omitting some necessary contexts):
+\\[1ex]
+$
+\begin{array}{ll}
+\textbf{Judgement Form} & \textbf{Pronunciation}\\
+\sigma \mapsto \sigma' & \text{Static expression $\sigma$ transitions to $\sigma'$.}\\
+\sigma~\mathtt{sval} & \text{Static expression $\sigma$ is a static value.}\\
+\sigma~\mathtt{styerr} & \text{Static expression $\sigma$ raises a type error.}
+\end{array}
+$
+\\[1ex]
+The iterated static transition judgement $\sigma \mapsto^{*} \sigma'$ is the reflexive, transitive closure of the static transition judgement, and the static evaluation judgement $\sigma \Downarrow \sigma'$ is derivable iff $\sigma \mapsto^{*} \sigma'$ and $\sigma'~\mathtt{sval}$.
+
+As suggested by the external type translation judgement above, external types are static values of a primitive sort written $\mathsf{Ty}$.
+We will return to how this works (and consider the metatheoretic implications of this choice) in Sec.
+\ref{sec:metamodules}.
+%One might notice the similarities between the specifications of the SL and the IL.
+In fact, the SL is a conservative extension of our IL.
+In practice, one would not implement them separately, but formally, it is more convenient to specify them separately.
 
@@
 Although metamodules are in some sense prior to TSMs, we will describe TSMs first, temporarily assuming that the EL does build in standard type structure, for pedagogical purposes.
-
-\section{Modularly Metaprogrammable Textual Syntax}\label{sec:syntax}
-Verse specifies a textual concrete syntax for the EL.\footnote{Although Wyvern specified a layout-sensitive concrete syntax, to avoid unnecessary distractions, we will specify a more conventional layout-insensitive concrete syntax for Verse.} %We have chosen to specify a layout-sensitive textual concrete syntax (i.e.
+\subsection{Module Language}
+The Verse module language is taken directly from Standard ML, with which we assume a working familiarity for the purposes of this proposal \cite{harper1997programming,MacQueen:1984:MSM:800055.802036} (a related module language, e.g.
+Ocaml's, would work just as well for our purposes).
+We will give examples of its use in Sections \ref{sec:syntax} and \ref{sec:metamodules}, but because it has been thoroughly studied in the literature, we will defer to prior work both here and in the dissertation for most formal details.
+In particular, we will first specify our contributions assuming a language without an ML-style module system, then make modifications to this specification to make it compatible with the addition of a module system.
+To do so, we will assume that modules are being tracked by a module context, without detailing how this context is populated.
+%This context will be analagous to the \emph{named type context} specified in the papers just cited.% Instead, we will consider it abstractly where it interacts with the core language.%\todo{reminder that we are not going to re-specify the module system}
+
+\section{Modularly Programmable Textual Syntax}\label{sec:syntax}
+To begin, let us assume a standard, ML-like semantics for the EL and consider its concrete syntax.
+Verse, like many major contemporary languages, specifies a textual concrete syntax for the EL.\footnote{Although Wyvern specified a layout-sensitive concrete syntax, to avoid unnecessary distractions, we will specify a more conventional layout-insensitive concrete syntax for Verse.} %We have chosen to specify a layout-sensitive textual concrete syntax (i.e.
 newlines and indentation are not ignored).
@@
 %For constructs that have an obvious analog in ML, we will omit a detailed explanation.
-Because the purpose of concrete syntax is to serve as a programmer-facing user interface, it is common practice to build in  derived syntactic forms (colloquially, \emph{syntactic sugar}) that capture common idioms more concisely or naturally.
+Because the purpose of concrete syntax is to serve as the programmer-facing user interface of the language, it is common practice to build in  derived syntactic forms (colloquially, \emph{syntactic sugar}) that capture common idioms more concisely or naturally.
 % (i.e.
 considering cognitive dimensions \cite{green1996usability}).
-For example, derived list syntax is built in to most functional languages, so that instead of having to write out \lstinline{Cons(1, Cons(2, Cons(3, Nil)))}, the programmer can equivalently write \lstinline{[1, 2, 3]}.
+For example, derived list syntax is built in to many functional languages, so that instead of having to write out \lstinline{Cons(1, Cons(2, Cons(3, Nil)))}, the programmer can equivalently write \lstinline{[1, 2, 3]}.
 Many languages go beyond this, building in derived syntax associated with various other types of data, like vectors (the SML/NJ dialect of SML), arrays (Ocaml), monadic commands (Haskell), syntax trees (Scala), XML documents (Scala, Ur/Web) and SQL queries (F\#, Ur/Web).
@@
 
-Verse takes a less \emph{ad hoc} and more modular approach -- rather than privileging  particular library constructs with primitive syntactic support, Verse exposes primitives that allow library providers to introduce new derived syntax of a variety of designs on their own.
+Verse takes a less \emph{ad hoc} approach -- rather than privileging  particular library constructs with primitive syntactic support, Verse exposes primitives that allow library providers to introduce new derived syntax of a variety of designs on their own, in a safe and modular manner.
 %Lists need no special consideration from the language specification.
@@
 %For forms with a clear analogy to a form in Standard ML, we will assume the  semantics are analagous without providing details.
-
-We will begin in Sec.
-\ref{sec:examples} by considering in detail a representative example for which such a mechanism might be useful: regular expression patterns expressed using abstract data types.
+We begin in Sec.
+\ref{sec:examples} by detailing a representative example for which such a mechanism would be useful: regular expression patterns expressed using abstract data types.
 In Sec.
-\ref{sec:syntax-existing}, we will demonstrate that the usual approach of using dynamic string parsing to introduce patterns is not ideal, and then survey existing alternatives, finding that they involve an unacceptable loss of modularity and other undesirable trade-offs.
+\ref{sec:syntax-existing}, we then demonstrate that the usual approach of using dynamic string parsing to introduce patterns is not ideal.
+We also survey existing alternatives to dynamic string parsing, finding that they involve an unacceptable loss of modularity and other undesirable trade-offs.
 In Sec.
-\ref{sec:syntax-contributions}, we outline our proposed mechanisms and discuss how they will resolve these issues.
+\ref{sec:syntax-contributions}, we outline our proposed alternatives -- \emph{typed syntax macros} and the related \emph{type-specific languages} -- and discuss how they resolve these issues.
+We also discuss how TSMs are specified in terms of the judgement forms introduced in the previous section.
 We conclude in Sec.
-\ref{sec:syntax-timeline} with a timeline for remaining work.
+\ref{sec:syntax-timeline} with a concrete timeline for the remaining work.
 
 \subsection{Motivating Example: Regular Expression Syntax}\label{sec:examples}
-Let us begin by taking the perspective of a regular expression library provider.
-Recall that regular expressions are a common way to capture patterns in strings \cite{Thompson:1968:PTR:363347.363387}.
-The abstract syntax of {patterns}, $p$, over strings, $s$, is specified as below:\[p ::= \textbf{empty} ~|~ \textbf{str}(s) ~|~ \textbf{seq}(p; p) ~|~ \textbf{or}(p; p) ~|~ \textbf{star}(p) ~|~ \textbf{group}(p)\]
+Let us begin by taking the perspective of a regular expression library provider. We assume the reader has some familiarity with regular expressions \cite{Thompson:1968:PTR:363347.363387}.
+
+\paragraph{Abstract Syntax} The abstract syntax of {patterns}, $p$, over strings, $s$, is specified as below:\[p ::= \textbf{empty} ~|~ \textbf{str}(s) ~|~ \textbf{seq}(p; p) ~|~ \textbf{or}(p; p) ~|~ \textbf{star}(p) ~|~ \textbf{group}(p)\]
 One way to express this abstract syntax is by defining a recursive sum type \cite{pfpl}.
@@
 casetype Pattern {
-    Empty
-  | Str of string
-  | Seq of Pattern * Pattern
-  | Or of Pattern * Pattern
-  | Star of Pattern
-  | Group of Pattern
+    Empty | Str of string | Seq of Pattern * Pattern 
+    | Or of Pattern * Pattern | Star of Pattern | Group of Pattern
 }
@@
 Indeed, there may be many ways to represent regular expression patterns, each with different performance trade-offs.
-For these reasons, a better approach in Verse, as in ML, is to abstract over the choice of representation using  the module system's support for type abstraction.
+For these reasons, a better approach in Verse, as in ML, is to abstract over the choice of representation using  the module system's support for generative type abstraction.
 In particular, we can define the following \emph{module signature}, where the type of patterns, \lstinline{t}, is held abstract:
@@
 \end{lstlisting}
- Any client of a module \lstinline{P} that satisfies \lstinline{PATTERN}, written \lstinline{P : PATTERN}, can manipulate patterns as terms of type \verb|P.t| using the interface described by this signature.
-By holding the representation type of patterns abstract, the burden of proving that the case analysis function cannot be used by clients to distinguish patterns with the same normal form is localized to each module implementing this signature.
-The details are standard and not particularly relevant for our purposes, so we omit them here.
-
-\paragraph{Concrete Syntax} The abstract syntax of patterns is too verbose to be used directly in all but the most trivial examples, so patterns are conventionally written using a more concise concrete syntax.
-For example, the concrete syntax \lstinline{A|T|G|C} corresponds to abstract syntax with the following much more verbose expression:
+ Client of any module \lstinline{P} that satisfies \lstinline{PATTERN}, written \lstinline{P : PATTERN}, manipulate patterns as values of the type \verb|P.t| using the interface described by this signature.
+The identity of this type is held abstract outside the module during typechecking (i.e.
+it acts as a newly generated type) so that the burden of proving that there is no way to use the case analysis function to distinguish patterns with the same normal form is local to the module.
+%The details are standard and not particularly relevant for our purposes, so we omit them here.
+
+\paragraph{Concrete Syntax} The abstract syntax of patterns is too verbose to be practical  in all but the most trivial examples, so programmers conventionally write patterns using a more concise concrete syntax.
+For example, the concrete syntax \lstinline{A|T|G|C} corresponds to the following much more verbose pattern expression:
 \begin{lstlisting}[numbers=none,mathescape=|]
@@
 \subsubsection{Dynamic String Parsing}
-To expose this more concise concrete syntax for regular expression patterns to clients, the most common approach is to provide a parse function that transforms strings to  patterns.
-Because, as just mentioned, there  may be many implementations of the \lstinline{PATTERN} interface, the standard approach is to define a parameterized module (a.k.a.
-\emph{functor} in SML) defining utility functions like this abstractly:
+To expose this more concise concrete syntax for regular expression patterns to clients, the most common approach is to provide a function that parses strings to produce patterns.
+Because, as just mentioned, there  may be many implementations of the \lstinline{PATTERN} signature, the usual approach is to define a parameterized module (a.k.a.
+a \emph{functor} in SML) defining utility functions like this abstractly:
 
@@
 module PatternUtil(P : PATTERN) => mod {
-  fun parse(s : string) : P.t = (* ... pattern parser here ... *)
-}
-\end{lstlisting}
-This allows the client of any module \lstinline{P : PATTERN} to use the following definitions:
+  fun parse(s : string) : P.t => (* ... pattern parser here ... *)
+}
+\end{lstlisting}
+This allows a client of any module \lstinline{P : PATTERN} to use the following definitions:
 \begin{lstlisting}[numbers=none]
@@
 \end{lstlisting}
-When compiling an expression like this, the programmer would see an error message like \verb|error: illegal escape character|.\footnote{This is the error message that \texttt{javac} produces.
-When compiling an analagous expression using SML of New Jersey (SML/NJ), we encounter the rather bizarre error message \texttt{Error: unclosed string}.} In a small lab study, we observed that this sort of error was common, and that diagnosing the problem nearly always required a non-trivial amount of time for even experienced programmers who had not used regular expressions recently \cite{Omar:2012:ACC:2337223.2337324}.
-One workaround with higher syntactic cost is to  double backslashes:
+When compiling an expression like this, the client would see an error message like \verb|error: illegal escape character|.\footnote{This is the error message that \texttt{javac} produces.
+When compiling an analagous expression using SML of New Jersey (SML/NJ), we encounter the rather bizarre error message \texttt{Error: unclosed string}.} In a small lab study, we observed that this class of error was common, and often temporarily confused even experienced programmers if they had not used regular expressions recently \cite{Omar:2012:ACC:2337223.2337324}.
+The standard workaround has higher syntactic cost -- one must double all backslashes:
 \begin{lstlisting}[numbers=none]
@@
 
-Some languages, anticipating such modes of use, build in alternative string forms that leave escape sequences uninterpretted. For example, Ocaml allows the following, which is perhaps more acceptable:
+Some languages, anticipating such modes of use, build in alternative string forms that leave escape sequences uninterpretted.
+For example, Ocaml supports the following, which has only a constant syntactic cost:
 \begin{lstlisting}[numbers=none]
@@
 \item The next problem is that pattern parsing does not occur until the pattern is evaluated.
-For example, the following malformed pattern will only trigger an error when this expression is evaluated during the full moon: %Achieving this goal is an explicit goal of this proposal, so we are obviously not happy with this.
+For example, the following malformed pattern will only trigger an exception when this expression is evaluated during the full moon: %Achieving this goal is an explicit goal of this proposal, so we are obviously not happy with this.
 
@@
 \end{lstlisting}
-Such problems can sometimes be discovered via testing, but empirical data gathered from large open source projects suggests that there remain many malformed regular expression patterns that are not detected by a project's test suite ``in the wild'' \cite{spishak2012type}.
-
-We might instead attempt to treat the well-formedness of patterns constructed from strings as a static verification condition.
-Automatically proving that this condition holds throughout a program is difficult to do in general, because it involves reasoning about arbitrary dynamic behavior.
-Here, the decision procedure must know that the variable \lstinline{pattern} is equal to the function \lstinline{PU.parse}.
-If the argument had not been written literally but rather computed, e.g.
-if we had written \lstinline{"SSTR(GESTR" ^ "SSTRCESTR"} where \lstinline{^} is the string concatenation function applied in infix style, we would need to be able to establish that this expression is equivalent to the string \lstinline{"SSTR(GCESTR"}.
-For patterns that are dynamically constructed based on input to a function (further discussed below), the problem is harder still.
-Asking the client to provide a proof of well-formedness would defeat the purpose of providing a concise concrete syntax.
-
-\item Parsing patterns dynamically also incurs a performance cost.
-Regular expression patterns are often used on large datasets, so it is easy to incur this cost repeatedly.
+In other words, this approach leaves the well-formedness of patterns constructed from strings as a static verification condition. 
+Though violations of this condition can sometimes be discovered dynamically via testing, empirical data gathered from large open source projects suggests that there remain many malformed regular expression patterns that are not detected by a project's test suite ``in the wild'' \cite{spishak2012type}.
+
+Statically verifying that this condition holds throughout a program is quite difficult to automate in general, because it requires reasoning about arbitrary dynamic behavior.
+In this example, a decision procedure would need to be able to establish that the variable \lstinline{pattern} is equal to the parse function \lstinline{PU.parse}.
+If the string argument had not been written literally but rather computed, e.g.
+as \lstinline{"SSTR(GESTR" ^ "SSTRCESTR"} where \lstinline{^} is the string concatenation function applied in infix style, it would also need to be able to establish that this expression is equivalent to the string \lstinline{"SSTR(GCESTR"}.
+For patterns that are dynamically constructed based on input to a function (further discussed below), the problem is not solved by ``simply'' statically evaluating the expression.
+
+Of course, asking the client to provide a proof of well-formedness would completely defeat the purpose -- providing a concise concrete syntax.
+
+\item Dynamic string parsing also necessarily incurs dynamic cost.
+Regular expression patterns are common when processing large datasets, so it is easy to inadvertently incur this cost repeatedly.
 For example, consider mapping over a list of strings:
@@
 \end{lstlisting}
-To avoid incurring the parsing cost for each element of \lstinline{exmpl_list}, the programmer or compiler must move the parsing step out of the transformation or ensure that an appropriately tuned caching strategy is being used.\footnote{Anecdotally, based on my experience in bioinformatics, the most common situation is that the compiler is not generally willing to partially evaluate expensive computations like pattern parsing out of such expressions automatically, and the programmer is often unaware of the performance cost.} %If the programmer does it, it can sometimes make the program more difficult to read.
+To avoid incurring the parsing cost for each element of \lstinline{exmpl_list}, the programmer or compiler must move the parsing step out of the closure or, barring that, ensure that an appropriately tuned memoization (i.e.
+caching) strategy is being used.\footnote{Anecdotally, in scientific code using standard contemporary compilers, this optimization is not automatic.
+Compilation caches are more often provided by the regular expression library, but like any cache, it is difficult to reason compositionally about performance in their presence.} %If the programmer does it, it can sometimes make the program more difficult to read.
 
@@
 
-\item The next problem is that dynamic string parsing only decreases the cost of statically known patterns, but not those constructed using other values.
+\item The next problem is that dynamic string parsing mainly decreases the syntactic cost of statically known patterns.
+Patterns constructed dynamically cannot easily benefit from this technique.
 For example, consider this function from strings to patterns:
@@
 We needed to use both dynamic string parsing and explicit applications of pattern constructors to achieve the intended semantics.
-It is difficult to capture idioms like this more concisely using dynamic string parsing (we will see more concise syntax below).
-
-\item Finally, for functions like the one just described where we are constructing patterns on the basis of data of type \lstinline{string}, it is easy for programmers to make  mistakes  using string concatenation.
-For example, consider the following ``alternative'' definition of \lstinline{example}:
+It is difficult to capture idioms like this more concisely using dynamic string parsing (we will see an example of syntax that does capture such idioms below).
+
+\item Finally, for functions like \lstinline{example} where we are constructing patterns on the basis of data of type \lstinline{string}, using strings coincidentally to introduce patterns makes it easy for programmers to use string concatenation in subtly incorrect ways.
+For example, consider the following seemingly more readable definition of \lstinline{example}:
 \begin{lstlisting}[numbers=none,escapechar=|]
@@
 
-Note that both functions have the same type and behave identically at many inputs, particularly those that would be expected during typical executions of the program (i.e.
+Both \lstinline{example} and \lstinline{example_bad} have the same type and behave identically at many inputs, particularly those that would be expected during typical executions of the program (i.e.
 alphabetic names).
-It is only when \lstinline{name} contains special characters that have meaning in the concrete syntax of patterns that a problem arises.
-In applications that query sensitive data, problems like this lead to \emph{injection attacks}, which are common and can be catastrophic from the perspective of software security \cite{owasp2013}.
-Proving that mistakes like these have not been made involves reasoning about complex run-time data flows, so it is once again difficult to automate.
+It is only when the input \lstinline{name} contains special characters that have meaning in the concrete syntax of patterns that a problem arises.
+In applications that query sensitive data, mistakes like this lead to \emph{injection attacks}, which are among the most common and catastrophic security threats today \cite{owasp2013}.
+Proving that mistakes like these have not been made involves reasoning about complex run-time data flows, so it is once again notoriously difficult to automate.
  %Ultimately, of course, mistakes like this are the fault of a programmer using a flawed heuristic, and they could be avoided with discipline.
@@
 
-The difficulties above are not unique to regular expression patterns.
-Whenever a library encourages the use of dynamic string parsing to address the issue of syntactic cost (which is, fundamentally, not a dynamic issue), such problems can arise.
+The problems above are not unique to regular expression patterns.
+Whenever a library encourages the use of dynamic string parsing to address the issue of syntactic cost (which is, fundamentally, not a dynamic issue), these problems arise.
 %Strings are, simply put, not ideally suited for this task.
-	This has motivated much research on reducing the need for dynamic string parsing \cite{Bravenboer:2007:PIA:1289971.1289975}.
+	This fact has motivated much research on reducing the need for dynamic string parsing \cite{Bravenboer:2007:PIA:1289971.1289975}.
 Existing alternatives can be broadly classified as being based on either \emph{direct syntax extension} or \emph{static term rewriting}.
@@
 \subsubsection{Direct Syntax Extension}\label{sec:syntax-extension}
-One tempting approach is to use a system that gives library providers the power to directly extend the concrete syntax of our language with new derived syntactic forms.
+One tempting alternative to dynamic string parsing is to use a system that gives the users of a language the power to directly extend its concrete syntax with new derived forms.
 %for regular expression patterns.% for patterns.
@@
 
-Unfortunately, this approach is perilous because the systems mentioned thusfar cannot guarantee that \emph{syntactic conflicts} between such extensions will not arise.
+For patterns constructed dynamically, we can define \emph{splicing syntax} that allows us to write string and pattern expressions inline (distinguished by prefixes \lstinline{@} and \lstinline{%}, respectively).
+For example, we can write \lstinline{example} equivalently, but at much lower syntactic cost, like this:
+\begin{lstlisting}[numbers=none,escapechar=|]
+fun example_concise(name : string) => 
+  /@name: %ssn/
+\end{lstlisting}
+%The body of this function elaborates to the body of \lstinline{example_fixed} as shown above. 
+Had we mistakenly written \lstinline{%name}, we would encounter only a static type error, rather than the  silent injection  vulnerability discussed above.
+
+All of the problems of dynamic string parsing described above are solved by this approach.
+Unfortunately, it introduces new problems.
+First, the systems mentioned thusfar cannot guarantee that \emph{syntactic conflicts} between such extensions will not arise.
 As stated directly in the  Coq manual: ``mixing different symbolic notations in [the] same text may cause serious parsing ambiguity''.
@@
 
-Putting this aside, there is also the question of how newly introduced forms elaborate to forms in our base grammar.
+Putting this aside, we must also consider the question of how newly introduced forms elaborate to forms in our base grammar.
 In our example, this is tricky because we have defined a modular encoding of patterns -- which particular module should the elaboration use?
@@
 Such a \emph{hygiene discipline} is well-understood only when performing term-to-term rewriting (discussed below) or in simple language-integrated rewrite systems like those found in Coq.
-For mechanisms that operate strictly at the level of context-free grammars, this issue has not been addressed.
+For mechanisms that operate strictly at the level of context-free grammars, this issue is not easy to address.
 
@@
 \end{lstlisting}
-For patterns constructed compositionally, we can define \emph{splicing} syntax for both strings (using prefix \lstinline{@}) and other patterns (using prefix \lstinline{%}).
-For example, we can write \lstinline{example_bad} instead like this:
-\begin{lstlisting}[numbers=none,escapechar=|]
-fun example_fixed1(name : string) => 
-  edu_cmu_verse_rx_pattern P /@name: %ssn/
-\end{lstlisting}
-%The body of this function elaborates to the body of \lstinline{example_fixed} as shown above. 
-Had we mistakenly used the pattern splicing syntax, writing \lstinline{%name}, rather than the string splicing syntax, \lstinline{@name}, we would encounter only a static type error, rather than the  silent injection  vulnerability that we discussed earlier.
-
-A general problem with approaches like these (in addition to those already mentioned) is that they  suffer from a paucity of direct reasoning principles -- given an unfamiliar piece of syntax, there is no straightforward method for determining what type it will have, or even for identifying which extension determines its elaboration, causing difficulties for both humans (related to code comprehension) and tools.
+
+Another problem with the approach of direct syntax extension is that there is no typing discipline -- given an unfamiliar piece of syntax, there is no straightforward method for determining what type it will have, causing difficulties for both humans (related to code comprehension) and tools.
 
@@
 An alternative approach is to leave the concrete syntax of the language fixed, but repurpose it for novel ends using a \emph{local term-rewriting system}.
-The LISP macro system \cite{Hart63a} is the most prominent example of such a system.
-Na\"ively, this system also suffers from the problem of unhygienic variable capture, but  modifications of this system, notably in the Scheme dialect of LISP, support reasoning hygienically about the rewriting that a macro performs \cite{Kohlbecker86a}.
-In languages with a stronger static type discipline, variations on macros that restrict rewriting to a particular type and perform the rewriting statically have also been studied \cite{Herman10:Theory,ganz2001macros} and integrated into languages, e.g.
+The LISP macro system \cite{Hart63a} is perhaps the most prominent example of such a system.
+Early variants of this system suffered from the problem of unhygienic variable capture just described, but  later variants, notably in the Scheme dialect of LISP, brought support for enforcing hygiene \cite{Kohlbecker86a}.
+In languages with a stronger static type discipline, variants of macros that restrict rewriting to a particular type and perform the rewriting statically have also been studied \cite{Herman10:Theory,ganz2001macros} and integrated into languages, e.g.
 MacroML \cite{ganz2001macros} and Scala \cite{ScalaMacros2013}.
 
-The most immediate problem with using these in our example is that we are not aware of any such statically-typed macro system that integrates cleanly with an ML-style module system.
-However, let us imagine a macro system that would allow us to repurpose string syntax  as follows:
+The most immediate problem with using these for our example is that we are not aware of any such statically-typed macro system that integrates cleanly with an ML-style module system, i.e.
+macros cannot be parameterized by modules.
+However, let us imagine such a macro system.
+We could use it to repurpose string syntax  as follows:
 \begin{lstlisting}[numbers=none]
@@
 Here, \lstinline{pattern} is a macro parameterized by a module \lstinline{P : PATTERN}.
-It statically parses the provided string literal (which must still be written using an Ocaml-style literal here) to generate an elaboration, specified by the macro to be at type \lstinline{P.t}.
-
-For patterns that are constructed compositionally, we would need to get more creative.
-For example, we might repurpose the infix operators that the language normally uses for other purposes to support string and pattern splicing as follows:
+It statically parses the provided string literal (which must still be written using an Ocaml-style literal here) to generate an elaboration.
+The macro specifies that the elaboration will be of type \lstinline{P.t}.
+By parsing the string statically, we avoid the problems of dynamic string parsing for statically known patterns.
+
+For patterns that are constructed dynamically, we need to get a bit more creative.
+For example, we might repurpose the infix operators that are normally used for other purposes to support string and pattern splicing as follows:
 
@@
 
-While this does not leave us with syntax that is quite as clean as would be possible with a  na\"ive syntax extension,  it does avoid many of the problems with that approach: there cannot be syntactic conflicts (because the syntax is not extended), we can define macro abbreviations because macros are integrated into the language, there is a hygiene discipline that guarantees that the elaboration will not capture variables inadvertently, and by using a typed macro system, there is a typing discipline (i.e.
-we need not examine the elaboration directly to know what type it must have).
+While this does not leave us with syntax that is quite as distinct and concise as would be possible with a  na\"ive syntax extension,  it does avoid many of the problems with that approach: there cannot be syntactic conflicts (because the syntax is not extended at all), we can define macro abbreviations because macros are integrated into the language, there is a hygiene discipline that guarantees that the elaboration will not capture variables inadvertently, and using a typed macro system, there is a type discipline (i.e.
+programmers need not examine the elaboration to know what type a macro application must have).
 
@@
 \subsection{Contributions}\label{sec:syntax-contributions}
-We will now introduce a new primitive -- the \textbf{typed syntax macro} (TSM) -- that combines the syntactic flexibility of syntax extensions with the reasoning guarantees of statically-typed macros, and show how to integrate it with the Verse module system.
-This will address all of the problems brought up in the previous two subsections.
-We will introduce TSMs in Sec.
-\ref{sec:tsms}.
-
-We then go further in Sec.
-\ref{sec:tsls}, introducing \textbf{type-specific languages} (TSLs).
+Verse provides a new external primitive -- the \textbf{typed syntax macro} (TSM) -- that combines the syntactic flexibility of syntax extensions with the reasoning guarantees of statically-typed macros.
+TSMs can be parameterized by modules, so they can be used to define syntax valid at all modules satisfying a specified signature.
+As we will discuss in in Sec.
+\ref{sec:tsms} below, this addresses all of the problems brought up above, at moderate syntactic cost.
+
+To further decrease syntactic cost (which is, of course, the entire purpose of this effort), we go on in Sec.
+\ref{sec:tsls} to briefly introduce \textbf{type-specific languages} (TSLs).
 TSLs are TSMs associated directly with a type when it is generated.
-Verse leverages local type inference to implicitly control TSL dispatch, further decreasing syntactic cost (which is, of course, the entire purpose of this exercise).
-
-The result is that derived syntax defined by library providers using TSMs and TSLs provides comes at syntactic cost comparable to that of derived syntax built in primitively by a language designer or  using a na\"ive tool like Camlp4, without the associated problems.
+Verse leverages local type inference to implicitly control TSL dispatch.
+
+Syntax defined by library providers using TSMs and TSLs comes at syntactic cost comparable to that of derived syntax built in primitively by the language designer or  using a na\"ive syntax extension tool like Camlp4, without resulting in dialect formation.
+As such, these \emph{ad hoc} approaches can, in large part, be discarded.
+That said, we examine some limitations of our approach in Sec.
+\ref{sec:tsms-limitations}.
 
@@
 %A typed syntax macro is invoked by applying it to a \emph{delimited form}, which can contain  arbitrary syntax in its \emph{body}.  
-Let us consider the following concrete external expression:
+To introduce TSMs, let us consider the following concrete external expression:
 \begin{lstlisting}[numbers=none]
@@
 \end{lstlisting}
-Here, a \emph{parameterized TSM}, \lstinline{pattern}, is being applied first to a module parameter, \lstinline{P}, then to a \emph{delimited form}, \lstinline{/SURLA|T|G|CEURL/} (a number of alternative delimiters are also provided by Verse's concrete syntax and could equivalently be used).
+Here, a \emph{parameterized TSM}, \lstinline{pattern}, is being applied first to a module parameter, \lstinline{P}, then to a \emph{delimited form}, \lstinline{/SURLA|T|G|CEURL/}.
+Note that a number of alternative delimiters are also provided by Verse's concrete syntax and could equivalently be used.
 The TSM statically parses the \emph{body} of the provided delimited form, i.e.
@@
 %The original expression, above, is statically rewritten to this expression.
-The definition of the parameterized TSM \lstinline{pattern} has the following basic form:
+The definition of \lstinline{pattern} looks like this:
 \begin{lstlisting}[numbers=none]
 syntax pattern(P : PATTERN) at P.t {
-  fn (ps : ParseStream) => (* pattern parser here *)
-}
-\end{lstlisting}
-We first identify the TSM, then specify the module parameter.
-Note that identifying it as \lstinline{P} here is not important, i.e.
-the TSM can be used with \emph{any} module matching signature \lstinline{PATTERN}.
-The module parameter is used in the type annotation \lstinline{at P.t}, which specifies that all elaborations successfully generated by applying this TSM to a module parameter \lstinline{P} and a delimited form will be of type \lstinline{P.t}.
-Elaborations are generated by the static action of the parse function defined on the second line.
-This parse function must be  a static value of type \lstinline{ParseStream -> Exp}.
-Both of these types are defined in the Verse \emph{prelude}, which is a set of definitions available ambiently.
-The type \lstinline{ParseStream} will give the function access to the {body} of the delimited form (in blue above) and the type \lstinline{Exp} is a case type that encodes the abstract syntax of external expressions.
-In other words, this function parses the body of the delimited form to generate an encoding of the elaboration.
-
-The parse function can identify portions of the body of the delimited form that should be treated as spliced expressions, allowing us to support splicing syntax as described in Sec.
-\ref{sec:syntax-extension}:
+  static fn (ps :: ParseStream) :: Exp => (* pattern parser here *)
+}
+\end{lstlisting}
+This TSM definition first identifies the TSM as \lstinline{pattern}, then specifies a module parameter, \lstinline{P}, which must match the signature \lstinline{PATTERN}.
+Note that identifying the module parameter as \lstinline{P} here is an entirely local choice, i.e.
+the TSM can be applied to \emph{any} module parameter matching  \lstinline{PATTERN}.
+This parameter is used in the type annotation \lstinline{at P.t}, which specifies that all elaborations that arise from the application of this TSM to a module \lstinline{P} and a delimited form must necessarily be of type \lstinline{P.t}.
+These elaborations arise by the static action of the parse function defined next, within braces.
+It is written as a static function of sort \lstinline{ParseStream -> Exp}.
+Both of these are defined in the Verse \emph{prelude}, which is a set of definitions available ambiently.
+The sort \lstinline{ParseStream} will give the function access to the {body} of the delimited form (in blue above) and the sort \lstinline{Exp}  encodes the abstract syntax of external expressions.
+%So the parse function parses the body of the delimited form to generate an encoding of the elaboration.
+
+To support splicing syntax as described in Sec.
+\ref{sec:syntax-extension}, the parse function must be able to extract external subexpressions directly from the parse stream.
+For example, consider the client code below:
 \begin{lstlisting}[numbers=none]
@@
 \end{lstlisting}
-%For concision in analytic positions, i.e. those where an expected type is known due to, for example, a type ascription, the module parameter \lstinline{P} can be inferred:
-%\begin{lstlisting}[numbers=none]
-%let N : P.t = pattern /SURLA|T|G|CEURL/
-%\end{lstlisting}
-
-A hygiene mechanism ensures that only portions of the generated elaboration derived from such spliced expressions can refer to the variables at the use site, preventing inadvertent variable capture by the elaboration.
-Put another way, the  elaboration logic must be valid in any typing context.
-
-\paragraph{Formalization} To give a flavor of the formal specification underlying TSMs, let us look at the rule for handling TSM application in a synthetic position: 
+The subexpressions \lstinline{name} and \lstinline{ssn} on the last line occur directly in the parse stream.
+When the parse function encounters these, it asks the \lstinline{ParseStream} to extract these as \emph{spliced expressions} for use in the elaboration.
+For example, the elaboration generated for the body of \lstinline{example_tsm} above would, if written concretely with spliced expressions marked, be:
+\begin{lstlisting}[numbers=none]
+P.Seq(P.Str(<spliced>(name)), P.Seq(P.Str ": ", <spliced>(ssn)))
+\end{lstlisting}
+The hygiene mechanism then ensures that only these marked portions of the generated elaboration can refer to the variables at the use site, preventing inadvertent variable capture by the elaboration.
+For example, the following would not be a valid elaboration, because it has free variables that are not inside spliced subexpressions:
+\begin{lstlisting}[numbers=none]
+P.Seq(P.Str(name), P.Seq(P.Str ": ", ssn))
+\end{lstlisting}
+%Put another way, the  elaboration logic must be valid in any context.
+
+\paragraph{Formalization} To briefly introduce the formal specification of TSMs, let us consider the rule in the EL's semantics for handling TSM application to a delimited form.
+To simplify matters, we will first consider only unparameterized TSMs (i.e.
+TSMs defined at a single type, e.g.
+\lstinline{Pattern}, rather than a parameterized family of types):
 \begin{mathpar}
 \inferrule[syn-aptsm]{
-    \Delta \vdash s ~@~ \sigma~\{\iota_\text{parser}\}\\
-    \mathbf{parsestream}(body) = \iota_\text{ps}\\
-    \iota_\text{parser}(\iota_\text{ps}) \Downarrow \iota_\text{elab}\\
-    \iota_\text{elab} \uparrow e_\text{elab}\\\\
-    \Delta; \Gamma; \emptyset; \emptyset \vdash e_\text{elab} \Leftarrow \sigma \leadsto \iota
+    \Delta \vdash \psi ~@~ \sigma_\text{ty}~\{\sigma_\text{parser}\}\\
+    \mathbf{parsestream}(b) = \sigma_\text{ps}\\
+    \sigma_\text{parser}(\sigma_\text{ps}) \Downarrow \sigma_\text{elab}\\
+    \sigma_\text{elab} \uparrow e_\text{elab}\\\\
+    \Delta; \Gamma; \emptyset; \emptyset \vdash e_\text{elab} \Leftarrow \sigma_\text{ty} \leadsto \iota_\text{trans}
 }{
-    \Delta_\text{out}; \Gamma_\text{out}; \Delta; \Gamma \vdash \mathbf{aptsm}(s; body) \Rightarrow \sigma \leadsto \iota
+    \Delta_\text{out}; \Gamma_\text{out}; \Delta; \Gamma \vdash \mathbf{aptsm}(\psi; b) \Rightarrow \sigma_\text{ty} \leadsto \iota_\text{trans}
 }
 \end{mathpar}
-Note that the judgement forms are  slightly simplified here, e.g.
-they omit various contexts relevant only to other Verse primitives.
+Looking at the conclusion of the rule, we see that TSM application takes the abstract form $\mathbf{aptsm}(\psi; b)$.
+The metavariable $\psi$ ranges over \emph{TSM expressions} and $b$ ranges over {bodies} of delimited forms.
+TSM expressions consist only of TSM definitions here (cf.
+the definition of \lstinline{pattern} above), though when we formalize parameterized TSMs, we will distinguish TSM definitions from TSM expressions.%Note that for simplicity in the formal system, we have tacitly assumed that TSM names are drawn from a set that is disjoint from the set of variable names, so that function application and TSM application desugar to the distinct abstract forms.
+%In practice, we would use a common set of names and include a disambiguation step, but making the assumption of disjoint names allows us to avoid specifying this.
+%In practice, there would be an additional step that determined whether a function or a TSM was being applied based on the current context,
+
+Note also that we omit various contexts relevant only to other Verse primitives.
 We include only \emph{typing contexts}, $\Gamma$, which map variables to types, and \emph{type abstraction contexts}, $\Delta$, which track universally quantified type variables, both in the standard way.
-The premises have the following meanings:
+We describe why there are two of each -- the \emph{outer contexts} and the \emph{current contexts} -- below.
+
+The premises can be understood as follows, in order:
 \begin{enumerate}
-\item The first premise can be pronounced ``TSM expression $s$ is a TSM at type $\sigma$ under $\Delta$ with parser $\iota_\text{parser}$''.
-This judgement would handle the mechanisms of module parameter application, but we must elide the details here (a complete specification of this judgement represents the main piece of remaining work).
+\item The first premise can be pronounced ``TSM expression $\psi$ defines a valid TSM at type $\sigma_\text{ty}$ under $\Delta$ with parse function $\sigma_\text{parser}$''.
+The ``outputs'' of this judgement are the type $\sigma_\text{ty}$ (of sort $\mathsf{Ty}$) and the static parse function, $\sigma_\text{parser}$ (of sort $\mathsf{ParseStream} \rightarrow \mathsf{Exp}$), defined by the TSM.% We must omit the rules governing this judgement here (a complete specification of this judgement represents the main piece of remaining work, see below).
 % Application of a parameterized TSM to its parameters is handled by this judgement (not shown).
-\item The second premise creates a parse stream, $\iota_\text{ps}$, from the body of the delimited form.
-\item The third premise applies the parser to the parse stream to generate an encoding of the elaboration, $\iota_\text{elab}$. This is where the metaprogramming occurs.
-\item The fourth premise decodes $\iota_\text{elab}$, producing the elaboration $e_\text{elab}$.
-\item The fifth premise validates the elaboration by analyzing it against the type $\sigma$ under empty contexts.
-The \emph{current} typing and type abstraction contexts are ``saved'' for use when a spliced term is encountered during this process by setting them as the new \emph{outer contexts}.
-
-The outer contexts are switched back in only when encountering a spliced expression, which is marked:
+\item The second premise creates a parse stream, $\sigma_\text{ps}$, of sort $\mathsf{ParseStream}$, from the body of the delimited form.
+\item The third premise applies the parse function to the parse stream to generate an encoding of the elaboration, $\sigma_\text{elab}$, of sort $\mathsf{Exp}$.% Note that this is ``metaprogramming'' -- a static expression in the external semantics to generate an encoding of another external term.
+\item The fourth premise decodes $\sigma_\text{elab}$, producing the elaboration itself, $e_\text{elab}$.
+\item The final premise \emph{validates} the elaboration by analyzing it against the type $\sigma_\text{ty}$ specified by the TSM.
+The {current} typing and type abstraction contexts are ``saved'' for use when a spliced subexpression is encountered during this process by setting them as the new \emph{outer contexts}, as indicated.
+
+Variables in the outer contexts are not directly available to the elaboration, e.g. they are ignored by the (syn-var) rule:
+\begin{mathpar}
+\inferrule[syn-var]{ }{
+    \Delta_\text{out}; \Gamma_\text{out}; \Delta; \Gamma, x : \sigma \vdash x \Rightarrow \sigma \leadsto x
+}
+\end{mathpar}
+
+Outer contexts are switched back in when encountering marked spliced expressions (which arise as explained previously from spliced expressions in the parse stream):
 \begin{mathpar}
@@
 \end{mathpar}
-For example, the elaboration generated in \lstinline{example_tsm} above would, if written concretely, be:
-\begin{lstlisting}[numbers=none]
-P.Seq(P.Str(spliced(name)), P.Seq(P.Str ": ", spliced(ssn)))
-\end{lstlisting}
-
-Variables in the outer contexts cannot be used directly by the elaboration, e.g. they are ignored by the (syn-var) rule:
-\begin{mathpar}
-\inferrule[syn-var]{ }{
-    \Delta_\text{out}; \Gamma_\text{out}; \Delta; \Gamma, x : \sigma \vdash x \Rightarrow \sigma \leadsto x
-}
-\end{mathpar}
-
+
+This premise of (syn-aptsm) thus enforces both type safety and hygiene.
+
+%The translation of the elaboration becomes the translation in the conclusion of the rule.
 \end{enumerate}
@@
 \subsubsection{Type-Specific Languages (TSLs)}\label{sec:tsls}
-To further lower the syntactic cost of using TSMs, Verse also  supports \emph{type-specific languages} (TSLs), which allow library providers to associate a TSM directly with a generated type.
+With TSMs, we achieved the conventional syntax for regular expression patterns within delimiters, but we still explicitly had to apply the TSM and provide the module parameter each time.
+To further lower the syntactic cost of using TSMs, Verse also  supports \emph{type-specific languages} (TSLs).
+This allows library providers to associate a TSM directly with a generated type or parameterized type constructor.
 For example, a module \lstinline{P} can associate \lstinline{pattern} with \lstinline{P.t} as follows:
@@
 
- Local type inference then determines which TSM is applied when analyzing a delimited form not prefixed by a TSM name. For example, this is equivalent to \lstinline{example_tsm}:
+ Local type inference then determines which TSM is implicitly applied when analyzing a delimited form not prefixed by a TSM name.
+For example, this function is equivalent to \lstinline{example_tsm} (note the return type annotation):
 \begin{lstlisting}[numbers=none]
@@
 
-As another example, we can use TSLs to express derived list syntax. For example, if we use a case type, we can declare the TSL directly upon declaration as follows:
-\begin{lstlisting}[numbers=none]
-casetype list('a) {
-  Nil
-| Cons of 'a * list('a)
-} with syntax {
-  fn (body : ParseStream) => (* ... comma-delimited spliced exps ... *)
-}
-\end{lstlisting}
-Together, this allows us to write a list of patterns like this:
+As another example, we can use TSLs to express derived list syntax. For example, if we use a case type, we can define the TSL directly upon declaring the list type:
+\begin{lstlisting}[numbers=none]
+casetype list('a) { Nil | Cons of 'a * list('a) } with syntax {
+  static fn (body :: ParseStream) => 
+    (* ... comma-delimited spliced exps ... *)
+}
+\end{lstlisting}
+Together with the TSL for patterns, this allows us to write a list of patterns like this:
 \begin{lstlisting}[numbers=none]
@@
 However, we did not need to build in this syntax primitively.%The only constraint is that this syntax must be used in an analytic position, which we argue is actually better for code compren when encountering unfamiliar syntax.
-\subsection{Timeline}\label{sec:syntax-timeline}
-We have described and given a more detailed formal specification of TSMs in a recently published paper \cite{sac15}, and TSLs in a paper last year \cite{TSLs}, both in the context of the Wyvern language.
-The basic mechanics of elaboration encoding/decoding and hygienic elaboration validation have all been satisfyingly specified in this context.
-
-Wyvern does not specify an ML-style module system (type declarations provide the only form of type generativity), so in the context of Verse, there are some extensions that need to be defined in the formalization to handle module parameters.
-We plan to complete this in the course of writing the dissertation, and anticipate it requiring about 3 weeks of effort over Summer 2015.
-We emphasize again that we do not plan to specify the full module system in detail; we will assume that judgements of an appropriate form have been specified and detail only those portions that interface with the novel aspects of the core language.
+
+\subsubsection{Limitations}\label{sec:tsms-limitations}
+%\todo{revise this section}The main limitation of this approach is that it 
+Elaboration validation ensures that an incorrect parser cannot threaten type safety, but proving type safety inductively is subtle because the elaboration is not a subexpression of the expression it will replace.
+There is a ``simple'' solution to this problem: the elaboration cannot itself use TSMs or TSLs.
+This would be somewhat awkward in practice, so we plan to briefly discuss other more flexible solutions as well.
+
+Another concern is that if we define the SL such that it supports general recursion, this then causes typechecking to become undecidable, because the parse function may not terminate.
+In practice, this too should only be a minor concern -- a compiler might be expected to limit the running time of the parse function to ensure that compilation does not ``hang'' due to an incorrect parser.
+Related concerns would arise if the static language supported state or access to other external effects.
+Existing techniques for controlling effects would handle this problem, but this is not something we plan to discuss in detail here because our SL does not specify such effects.
+
+Another limitation is that TSMs as we have described them only capture idioms that occur within a single parameterized family of types,  not idioms that might span many (or all) types, e.g.
+control flow idioms.
+In fact, this is intentional, as it makes maintaining a type discipline in the presence of TSMs substantially easier -- neither humans nor tools need to examine the elaboration to determine what type it must necessarily have.
+We will, however, discuss other points in the design space in the dissertation.
+
+Finally, we have discussed expression syntax here, but have not explicitly considered pattern matching.
+In fact, we conjecture that it is completely straightforward to extend this work to allow a TSM to define both an expression parser and a pattern parser.
+We plan to concretely discuss this as the most immediate avenue for future work, but will not consider pattern matching in formal detail.
+
+\subsection{Timeline and Milestones}\label{sec:syntax-timeline}
+We have described and given a more detailed formal specification of TSMs in a recently published paper \cite{sac15}, and TSLs in a paper published last year \cite{TSLs}, both in the context of the Wyvern language.
+The mechanics of statically invoking the parse function, elaboration encoding/decoding and hygienic elaboration validation have all been detailed there.
+In the dissertation, I plan to first present the formal system closely following this presentation, updating it only to ensure that it is consistent with the work in the next section to present a unified design.
+The rule (syn-aptsm) above represents the most substantial step of this plan.
+I anticipate that this will take about \textbf{1.5 weeks} of further work.
+
+Wyvern does not have an ML-style module system (type declarations are the only form of type generation) and at the time, it did not have parameterized types, so to actually support the examples we have given in this section, we also need to update the formalization to handle parameterized TSMs.
+We plan to flesh this out in the course of writing the dissertation, and I anticipate it requiring 5.5 weeks of my time after the proposal is complete.
+We propose the following concrete milestones:
+
+\begin{enumerate}
+\item equip the EL with an appropriately structured module context \textbf{(2 weeks)}
+\item formally specify the syntax of parameterized TSM expressions \textbf{(0.5 weeks)}
+\item formally specify an updated version of the TSM expression validation judgement that appeared as the first premise of (syn-aptsm) above \textbf{(1 week)}
+\item update the metatheory established in the work cited above to handle this \textbf{(2 weeks)}
+\end{enumerate}
+
+To validate our claims of expressiveness, we plan to give several more examples drawn from existing languages.
+We will include those we have worked out in the papers just cited as well as additional examples that highlight the benefits of integration with Verse's ML-based module system.
+For example, we will show how to implement Haskell's primitive ``do'' syntax using TSLs together with a modular encoding of monads (as described on Prof.
+Harper's blog \cite{SML/Monads}).
+We will also briefly summarize the corpus analysis we conducted in \cite{TSLs} where we found a large number of  situations in large publicly available Java libraries where dynamic string parsing was apparently being used to control syntactic cost.
+We anticipate that writing these examples up will take an additional \textbf{1 week} of work, bringing the total for this section up to 8 weeks.
+%Adding in the time I will need to propose and some buffer time for unanticipated issues, my goal is to have this portion of the dissertation complete by the end of Summer 2015.
 
@@
 
-\section{Modularly Metaprogrammable Type Structure}\label{sec:metamodules}
-Let us now turn our attention to the type structure of the Verse EL.
-In contrast to other languages, which build in constructs like case types (shown briefly in use in the previous section), tuples, records, objects and so on primitively, in Verse these can all be expressed as modes of use of \emph{metamodules} (which are distinct from, though conceptually related to, {modules}, as we will see).
+\section{Modularly Programmable Type Structure}\label{sec:metamodules}
+In the previous section, we assumed that the EL had an otherwise standard, ML-like type structure and considered only TSMs and TSLs as departures from ML.
+In fact, constructs like case types (shown briefly in use in the previous section), tuples, records and objects are not built primitively into Verse.
+Instead, they can be expressed using \emph{metamodules} (which are distinct from, though closely related to, {modules}, as we will discuss below).
 Only polymorphic function types are built primitively into the Verse EL.
 
-This section is organized much like Sec.
+This section is organized like Sec.
 \ref{sec:syntax}.
-We begin with examples of type structure that we wish to express in Sec.
+We will begin with examples of type structure that we might want to express in Sec.
 \ref{sec:metamodules-example}, then discuss how existing approaches are insufficient in Sec.
 \ref{sec:metamodules-related}.
-Next, we outline how metamodules (together with TSMs and TSLs) serve to address these problems in Sec.
+Next, we outline how \emph{metamodules} (together with TSMs and TSLs) serve to address these problems in Sec.
 \ref{sec:metamodules-solution} and conclude with a timeline in Sec.
@@
 \subsection{Motivating Example: Labeled Tuples and Regular Strings}\label{sec:metamodules-example}
-As a simple introductory example, let us consider a type classifying conference papers:
-\begin{lstlisting}[numbers=none]
-let type Paper = ltuplex {SPCT
+As a simple introductory example, let us define a \emph{labeled tuple type} for conference papers:
+\begin{lstlisting}[numbers=none]
+let type Paper = ltuple {SPCT
    title : EPCTrstring /SURL.+EURL/
@@
 \end{lstlisting}
-The \lstinline{let type} construction defines a synonym, \lstinline{Paper}, for a type constructed by applying the \emph{type constructor} (or \emph{tycon}) \lstinline{ltuplex} to a statically valued ordered finite mapping from labels to types, here written using a conventional concrete syntax within curly braces (this is a mode of use of TSLs, as we will see).
-Its first component is labeled \lstinline{title} and specifies the regular string type \lstinline{rstring /SURL.+EURL/}.
-Regular string types are constructed by applying the tycon \lstinline{rstring} to a regular expression pattern, again written using a conventional concrete syntax.
-Regular string types classify values that behave like strings in the corresponding regular language, here the language of non-empty strings.
-The second component of type \lstinline{Paper} is labeled \lstinline{conf} and specifies a regular string type with two parenthesized groups, corresponding to a conference abbreviation and year.
+The \lstinline{let type} construction defines a synonym, \lstinline{Paper}, for a type constructed by applying the \emph{type constructor} (or \emph{tycon}) \lstinline{ltuple} to a \emph{component specification}, which is a  statically valued ordered finite mapping from labels to types, written using conventional concrete syntax.
+
+The first component is labeled \lstinline{title} and specifies the \emph{regular string type} \lstinline{rstring /SURL.+EURL/}.
+Regular string types are constructed by applying the tycon \lstinline{rstring} to a statically valued regular expression pattern, again written here using conventional concrete syntax like that discussed in the previous section.
+Regular string types classify expressions that behave like strings in the corresponding regular language, here the language of non-empty strings.
+The second component of type \lstinline{Paper} is labeled \lstinline{conf} and specifies a regular string type with two {parenthesized groups}, corresponding to a conference abbreviation and year.
 
@@
 \begin{lstlisting}[numbers=none]
-let val exmpl : Paper = {SCSSconf=>ECSS"SUSEXMPL 2015EUS"SCSS, title=>ECSS"SUSAn Example PaperEUS"SCSS}
+let val exmpl : Paper = {SCSSconf=>ECSS"SUSEXMPL 2015EUS"SCSS, title=>ECSS"SUSAn Example PaperEUS"}
 \end{lstlisting}
@@
 \end{lstlisting}
-Here, \lstinline{#} identifies an \emph{operator constructor} (or \emph{opcon}) parameterized by a static label, written literally here as \lstinline{<SNATconfENAT>}.
-The resulting \emph{operator}, \lstinline{# <SNATconfENAT>}, takes one \emph{argument}, \lstinline{exmpl}.
-
-Given a value of regular string type like this, we can project out captured groups using the \lstinline{#group} operator constructor, which is parameterized by a natural number referring to the group index:
+Here, \lstinline{#} identifies an \emph{operator constructor} (or \emph{opcon}) parameterized by a statically valued label, written literally here as \lstinline{<SNATconfENAT>}.
+The resulting \emph{operator}, \lstinline{# <SNATconfENAT>}, is then passed one \emph{argument}, \lstinline{exmpl}.
+An important point is that this operator is not a function, i.e.
+it cannot be given function type, because it is able to operate on values of \emph{any} labeled tuple type that has a component labeled \lstinline{conf}, not just \lstinline{Paper}.
+That type's component specification determines the type of the operation as a whole, here \lstinline{rstring /SURL([A-Z]+) (\d\d\d\d)EURL/}.
+
+We can project out the first captured group from the regular string \lstinline{exmpl_conf} using the \lstinline{#group} operator constructor, which is parameterized by a statically valued natural number referring to the group index:
 \begin{lstlisting}[numbers=none]
@@
 
-Labeled tuples support a number of other basic operations.
-For example, labeled tuples can be concatenated (with any common components updated with the value on the right) using the \lstinline{ltuple+} operator:
-\begin{lstlisting}[numbers=none]
-let val a : ltuplex {SPCTauthors : EPCTlist(rstring /SURL.+EURL/)} = {["SUSHarry Q. BovikEUS"]}
-let val exmpl_paper_final = ltuple+ exmpl_paper a
+We will equip labeled tuple types with an extended set of operators that go beyond the introduction and projection operators just discussed.
+For example, two labeled tuples can be concatenated (with common components updated with the value on the right) using the \lstinline{ltuple+} operator.
+Using this operator, we can add a component labeled \lstinline{authors} to \lstinline{exmpl}:
+\begin{lstlisting}[numbers=none]
+let val a : ltuple {SPCTauthors : EPCTlist(rstring /SURL.+EURL/)} = {["SUSHarry Q. BovikEUS"]}
+let val exmpl_paper_final = ltuple+ exmpl a
 \end{lstlisting}
@@
 \end{lstlisting}
+There are many other operators that we might wish to describe, for both labeled tuples and regular strings, but for our purposes it suffices to stop here.
 
 \subsection{Existing Approaches}\label{sec:metamodules-related}
-Before considering labeled tuples and regular strings as a mode of use of metamodules, let us consider some alternative approaches.
-
-\subsubsection{Labeled Tuple Types}
-Recall that Verse builds in only nullary and binary products into its internal language, so the type structure described above for labeled tuples  is not directly expressible using IL primitives.
-The EL does not build in even these.
-However, so as to separate concerns, let us assume for the moment that regular string types are available in the EL: %So to consider existing approaches, we have to at least somewhat weaken our insistence on minimalism.
+Before describing how we express the type structure of labeled tuples and regular strings in Verse using metamodules, let us consider some alternative approaches for each.
+
+\subsubsection{Labeled Tuples}
+Recall that Verse builds only nullary and binary products into the IL, so we cannot express the type structure described above for labeled tuples  directly using IL primitives.
+The Verse EL does not build in even these, but so as to separate concerns, let us assume for the moment that regular string types are available in the EL: %So to consider existing approaches, we have to at least somewhat weaken our insistence on minimalism.
 \begin{lstlisting}[numbers=none]
@@
 \paragraph{Modules}
-One approach we might take is to attempt to express the type structure of any particular labeled tuple type using the module system.
+It is reasonable to ask if it is possible to express the type structure of any particular labeled tuple type using the module system.
 %For simplicity below, let us slightly weaken our insistence on minimalism and build binary products into the EL (and, for now, assume that regular string types are also built in in some suitable manner).%This is not strictly necessary -- we could use a Church-style encoding -- but this simplifies our examples.%then we could try to use the module system to express the type structure of any particular labeled tuple type.
 For our example above:
-For our example type \lstinline{Paper} from above, we start by defining the following signature:
+For our example type \lstinline{Paper} from above, we might start by defining the following signature:
 \begin{lstlisting}
@@
 \end{lstlisting}
-We've simply taken all possible valid introductory and projection operators and called for their expression as  functions, moving label parameters into identifiers.
-Even with our minimal EL, we could  implement the semantics of these operators against this signature using a Church-style encoding (the details are standard and omitted).
-Alternatively, if we slightly relax  our insistence on minimalism and add binary products to the EL, we could use them to implement these operators as well:
+Here, we've taken all possible valid introductory and projection operators and specified their semantics using function types.
+Expressing labels explicitly is an important facet of having labeled tuple types in a language, so we put them into the function identifiers.
+Even with our minimal EL, we could  implement the semantics of these operators against this signature using a Church-style encoding (the details are standard and omitted here).
+Alternatively, if we slightly relax  our insistence on minimalism and add binary products to the EL, we could use them to implement our desired semantics as well, like this:
 \begin{lstlisting}
@@
 
-However, there are several problems with this basic approach.
+There are several fundamental problems with this approach.
 First, not every module matching the signature \lstinline{PAPER} correctly implements the semantics of the operators we seek to express, so each such module would itself need to be verified.
-For example, we need to verify the universal condition that for all \lstinline{e : Paper.t} and \lstinline{f : Paper.t -> T}, we have that \lstinline{f(e)} is equivalent to \lstinline{f(Paper.intro (Paper.prj_title e) (Paper.prj_conf e))}.
+In particular, we need to verify the universal condition that for all \lstinline{e : Paper.t} and \lstinline{f : Paper.t -> T}, we have that \lstinline{f(e)} is equivalent to \lstinline{f(Paper.intro (Paper.prj_title e) (Paper.prj_conf e))}.
 The volume of boilerplate code that needs to be generated and verified manually is factorial in the number of components of the labeled tuple type we are expressing (because we must encode each permutation of label orderings with a distinct introductory function).% Not all modules satisfying the signature above correctly implement the semantics, so this boilerplate must be checked for correctness in some way.
 
-Another issue is that we can only reasonably express the introductory and projection operators by enumerating them in this way.
-To enumerate all possible operators that arise from the opcon \lstinline{ltuple-} as functions would require constructing an encoding of every labeled tuple type that might arise from dropping one or more components.
-This would add an  exponential factor to our volume of boilerplate code.
-Finally, there is simply no way to define \lstinline{ltuple+} as a finite collection of functions because there are infinitely many extensions of any labeled tuple type.
+Even if we were to automate this (using, for example, \emph{type macros} as found in Scala \cite{ScalaMacros2013}, adapted to a language with modules), another issue is that we can only reasonably express the introductory and projection operators by specifying their semantics as functions like this.
+To enumerate all possible operators that arise from the opcon \lstinline{ltuple-} using functions would require constructing an encoding of every labeled tuple type that might arise from dropping one or more components from the labeled tuple in question.
+If fully enumerated, this would add an  exponential factor to our volume of boilerplate code.
+Finally, there is simply no way to specify the semantics \lstinline{ltuple+} with a finite collection of functions because there are infinitely many extensions of any labeled tuple type.
 % of this operator would need to be implemented by clients manually, defeating the purpose of the operator.
@@
 \paragraph{Records and Tuples}
-If, instead of settling for such a module-based encoding, we further weaken our insistence on minimalism and primitively build in record/tuple types, as SML and many other languages have done, let us consider what might be possible.
-We note at the outset that records and tuples too are library constructs in Verse, but because they are so commonly built in to other languages, this situation is worth considering.%we can retreat from trying to use the module system, but have only slightly better options.
-
-The simplest approach we might consider taking is to directly map each labeled tuple type to a record type having the same component mapping:
+If, instead of this module-based encoding, we further weaken our insistence on minimalism and primitively build record and tuple types into the EL, as SML and many other languages have done, let us consider what might be possible.
+Note that these too are library constructs in Verse, but because they are so commonly built in to other languages, this situation is worth considering.%we can retreat from trying to use the module system, but have only slightly better options.
+
+The simplest approach we might consider taking is to directly map each labeled tuple type to a record type having an identically written component specification:
 \begin{lstlisting}[numbers=none]
@@
 }\end{lstlisting}
-Unlike labeled tuple types, record types are identified only up to component reordering, so this mapping does not preserve certain type disequalities between labeled tuple types.
-This means it is not  possible to allow clients to omit component labels and rely on the positional information available in the type when introducing a value of this type.%, because this information is not maintained by the type system.
-
-To work around this limitation if we do not otherwise care about preserving these type disequalities, we could define two conversion functions involving unlabeled tuples:
+Unlike labeled tuple types, record types are identified up to component reordering, so this mapping does not preserve certain type disequalities between labeled tuple types.
+It is not  possible to allow clients to omit component labels and rely on the positional information available in the type when introducing a value of record type for this reason.%, because this information is not maintained by the type system.
+
+To work around this limitation if this is the only reason we care about preserving these type disequalities, we could define two conversion functions involving unlabeled tuples:
 \begin{lstlisting}[numbers=none]
@@
 For clients to be able to rely on this approach, we must extrinsically verify that the library provider implemented these functions correctly.
-Clients must also bear the syntactic and run-time costs of explicitly invoking these functions.
-We could define a TSL for every such type to reduce this cost to clients, though this is an additional cost to providers.
+Clients must also bear the syntactic and dynamic costs of explicitly invoking these functions.
+We could define a TSL for every such type to reduce this cost to clients, albeit at an additional cost to providers.
 %This also means that we cannot omit the labels and rely on positional information when introducing a value of a record type in an analytic position, as we did above with type \lstinline{Paper}.
@@
 
-If on the other hand we do care about preserving the type disequalities in question, the workaround is even less elegant.
+If, on the other hand, we do care about preserving all type disequalities between labeled tuple types, the workaround is even less elegant.
 We first need to define a record type with component labels tagged in some sufficiently unambiguous way by position, e.g.:
@@
 }\end{lstlisting}
-To avoid exposing this directly to clients, we need two auxiliary type definitions:
+To avoid exposing these labels directly to clients, we need two auxiliary type definitions:
 \begin{lstlisting}[numbers=none]
@@
 \end{lstlisting}
-Clients again bear the cost of applying these conversion functions ahead of every operation and providers again bear the burden of defining this boilerplate code (which again has volume linear in the number of components, albeit with a constant factor that is again not easy to dismiss).
-And once again, the provider must verify that these functions were implemented correctly.
+Clients once again bear the costs of applying these conversion functions ahead of every operation and providers again bear the burden of defining this boilerplate code (which has volume linear in the number of components, albeit with a constant factor that is again not easy to dismiss) and verifying that these functions were implemented correctly.
 %We are interested in all such pragmatic issues, so this is not an entirely satisfying solution.
@@
 
-Finally, note that neither of these workarounds provides us with any way to uniformly express the more interesting operations, like  concatenation or dropping a component, that we discussed earlier.
+As with the modular encoding of labeled tuples we attempted above, neither of these encodings using records and unlabeled tuples provides us with any way to uniformly express the more interesting operations, like \lstinline{ltuple+} or \lstinline{ltuple-}, that we discussed earlier.
 
 \subsubsection{Regular Strings}
-Let us now turn to regular string types.
-
-\paragraph{Run-Time Checks}
-The most common alternative to regular strings in existing languages is to instead use standard strings (which themselves may be expressed as lists or vectors of characters together with derived syntax), inserting run-time checks around operations to maintain the regular string invariant dynamically.
-This clearly does not express the static semantics of regular strings, and incurs syntactic and run-time cost.
+Let us now turn to regular string types. Note that we have specified the full semantics of regular string types in a recent workshop paper \cite{sanitation-psp14}.
+
+\paragraph{Dynamic Checks}
+The most common alternative to regular strings in existing languages is to simply use standard strings (which themselves may be expressed as lists or vectors of characters together with derived syntax) and insert dynamic checks around operations to maintain the regular string invariant.
+This clearly does not express the static semantics of regular strings, and incurs syntactic and dynamic cost.
 
 \paragraph{Type Refinements}
-We might instead try to express the static semantics of regular strings by using standard strings and moving the logic governing regular string types into an external system of \emph{type refinements}  \cite{Freeman91}.
+We might attempt to recover the static semantics of regular strings by moving the logic governing regular string types into an external system of \emph{type refinements} \cite{Freeman91}.
 %This treats membership of a string in a specified regular language exclusively as a static verification condition.
@@
 \end{lstlisting}
-One problem with this is that there is no way to express the group projection operator described above. We cannot define a function \lstinline{#group} that can be used like this:
+One immediate limitation of this approach is that there is now no way to express the group projection operator described above.
+We cannot define a function \lstinline{#group} that can be used like this:
 \begin{lstlisting}[numbers=none]
@@
 \end{lstlisting}
-The reason is that this function would need information that is  available only in the type refinement of each regular string, i.e.
-the positions of the captured groups.
-We would instead need to dynamically match the string against the regular expression explicitly:
+The reason is that group projection does not have a function-like semantics -- it needs access to information that here is  available only in the type refinement of each regular string, i.e.
+the definitions of the captured groups.
+To access a group, we would instead need to dynamically match the string against the regular expression explicitly:
 
@@
 \end{lstlisting}
-Ensuring that there will not be an out-of-bounds error here would again require more sophisticated extrinsic reasoning.
+This has higher syntactic and dynamic cost. Moreover, proving that there will not be an out-of-bounds error in expressions like this requires more sophisticated reasoning about dynamic behavior.
 %Refinement types are strictly a layer above the language, and thus cannot affect its dynamic semantics.
@@
 For example, the fact that some value of type \lstinline{string} can be given the refinement \lstinline{rstring /A|B/} cannot be justification to locally alter its representation to, for example, a single bit, both because it could later be passed into a function expecting a standard string and because there are many other possible refinements.
-Of course, for regular strings, there may be limited benefit (or even some cost) to controlling representation differentially, but in other situations, this could have significant performance ramifications.
-For example, it is quite common to justify the omission of a primitive type of ``non-negative integers representable in 32 bits'' by the fact that this is ``merely'' a refinement of (mathematical) integers, but in so doing the compiler loses the ability to use a machine representation more suitable to this particular static invariant.%Refinement types cannot be used for this purpose without the possibility of conflict.
+Another perhaps more useful optimization that is not possible is to represent regular strings that have captured groups in a manner that ensures that group projection has constant dynamic cost (by running the match once when the string is introduced and caching the group values).
+
+Put another way, type refinements are useful when one needs to make ``behavioral'' distinctions within a type, but they are not appropriate when new ``structural'' distinctions are needed.
+Another example where this is relevant is when a language designer considers omitting a primitive type of ``non-negative integers representable in 32 bits'' because this seems ``merely'' to be a refinement of (mathematical) integers.
+But by eliminating the structural distinction, the compiler loses the ability to use a machine representation more suitable to this particular static invariant.
+Of course, structural distinctions are sometimes inconvenient because they require defining and applying coercions between distinct types, but it is certainly not the case that a language can reasonably predict all possible structural distinctions that might be useful \emph{a priori}.
+%Refinement types cannot be used for this purpose without the possibility of conflict.
 %This is also the reason why one cannot viably do away with primitive record types by claiming that they are refinements of finite mappings with labels as keys.
@@
 
-\paragraph{Modules}
-To create type disequalities, we might consider again turning to the module system, defining a signature for each regular string type that we wish to work with.
+% \paragraph{Modules}
+% To create type disequalities, we might consider again turning to the module system, defining a signature for each regular string type that we wish to work with.
 Again, this requires a non-trivial amount of boilerplate code:
-\begin{lstlisting}[numbers=none]
-signature TITLE = sig {
-  type t
-  val intro : string -> t
-  val strcase : t -> (unit -> 'a) -> (string * string -> 'a) -> 'a
-}
-
-signature CONF = sig {
-  type t
-  val intro : string -> t
-  val strcase : t -> (unit -> 'a) -> (string * string -> 'a) -> 'a
-  val prj_group_0 : t -> string
-  val prj_group_1 : t -> string
-}
-\end{lstlisting}
-
-To ensure that applications of the introductory operators are correct statically, we would need to couple this with a system of type refinements.
-Moreover, this again reveals itself to be too limited a solution when we need to express other operators.
-For example, expressing concatenation uniformly using a single function or a finite set of functions is impossible, for the same reason as applied to \lstinline{ltuple+} above, i.e.
-there are infinitely many regular string types that might be involved.
+% \begin{lstlisting}[numbers=none]
+% signature TITLE = sig {
+%   type t
+%   val intro : string -> t
+%   val strcase : t -> (unit -> 'a) -> (string * string -> 'a) -> 'a
+% }
+
+% signature CONF_GROUP_1 = sig {
+% 	type t
+% 	val intro : string -> t
+% 	val strcase : t -> (unit -> 'a) -> (sting * strng -> 'a) -> 'a 
+% }
+
+% signature CONF = sig {
+%   type t
+%   val intro : string -> t
+%   val strcase : t -> (unit -> 'a) -> (string * string -> 'a) -> 'a
+%   structure Grp0 : CONF_GROUP_0
+%   structure Grp1 : CONF_GROUP_1
+%   val prj_group_0 : t -> Grp0.t
+%   val prj_group_1 : t -> Grp1.t
+% }
+
+% \end{lstlisting}
+
+% As discussed previously, we do not have a guarantee that all implementations of these signatures correctly capture the semantics we seek to express, so they must each be verified externally.
+Moreover, we must externally verify that every application  of the \lstinline{intro} operator is statically correct as well (i.e.
+that it won't raise an exception), perhaps with a system of type refinements.
+As with all language-external verification approaches, it is difficult to propagate this fact into a compiler in a manner that guarantees that redundant dynamic checks are eliminated.
+And finally, this approach again reveals itself to be too limited when we need to express operators that cannot be expressed as functions.
+For example, expressing regular string concatenation uniformly using a single function or a finite set of functions is impossible, because there are infinitely many regular string types that might be involved.
 
@@
 Verse introduces a \emph{metamodule system} that gives library providers more direct control over the semantics of the EL.
-Using the metamodule system, library providers can express the type structure of labeled tuples and regular strings more directly, as we will show below.
+Using the metamodule system, library providers can express all of the types and operators discussed above.
 
@@
 ML) module system is organized around \emph{signatures} and \emph{modules}, the Verse metamodule system is organized around \emph{metasignatures} and \emph{metamodules}.
-Briefly summarized:
-\begin{itemize}
-\item A \emph{metasignature} specifies a set of type and operator constructors (tycons and opcons). Each is parameterized by a specified \emph{kind} of \emph{static value}. 
-\item A \emph{metamodule}  defines static functions that govern the typechecking and  translation to the IL of each type and operator constructor specified in the  metasignature. 
-\end{itemize}
-
-Static values are values of the \emph{typed static language} (SL), which itself forms a total typed lambda calculus, with \emph{static expressions} $\sigma$ classified by \emph{kinds} $\kappa$.
+These are most easily introduced by example.
+We consider only labeled tuple types in this section (regular string types will follow analagously; we will discuss regular string types in more detail in the dissertation).
+%Briefly summarized:
+%\begin{itemize}
+%\item A \emph{metasignature} specifies a set of type and operator constructors (tycons and opcons). Each is parameterized by a specified \emph{kind} of \emph{static value}. 
+%\item A \emph{metamodule}  defines static functions that govern the typechecking and  translation to the IL of each type and operator constructor specified in the  metasignature. 
+%\end{itemize}
+
+%Static values are values of the \emph{typed static language} (SL), which itself forms a total typed lambda calculus, with \emph{static expressions} $\sigma$ classified by \emph{kinds} $\kappa$.
 Types are one kind of static value.
@@
 \subsubsection{Example: Labeled Tuple Types via Metamodules}
-\begin{figure}[p!]
-\begin{lstlisting}
-metasignature LTUPLEX = metasig {
-  ^\label{line:ltuplex-c-sig}^tycon c of LblTyOrdMap
-  ^\label{line:intro-unlabeled-sig}^ana opcon intro_unlabeled of unit
-  ^\label{line:intro-labeled-sig}^syn opcon intro_labeled of list(Lbl)
-  ^\label{line:prj-sig}^syn opcon # of Lbl
-  ^\label{line:conc-sig}^syn opcon + of unit
-  ^\label{line:drop-sig}^syn opcon - of Lbl
-}
-
-^\label{line:ltplx-tsm-start}^syntax ltplx(L :: LTUPLEX, param :: LblTyOrdMap) at L.c(param) {
-  fn ps => (* 
-    - elaborates {e_1, ..., e_n} to L.intro_unlabeled(e_1; ...; e_n)
-    - elaborates {lbl1 => e_1, ..., lbln => e_n} to 
-        L.intro_labeled [lbl1, ..., lbln] e_1 ... e_n
-  *)
-}^\label{ltplx-tsm-end}^
-
-metamodule Ltuplex :: LTUPLEX = metamod {
-  ^\label{line:c-start}^(* translate labeled tuple types to nested products *)
-  tycon c {
-  	(* type translation generator: *)
-  	fn (param :: LblTyOrdMap) => fold (list_of_ltom param)
-      `SCSSunitECSS`
-      (fn ((lbl, ty), cur_ty_trans) => `SCSStrans(ECSStySCSS) * %ECSScur_ty_trans`)
-  }^\label{line:c-end}^
-
-  ^\label{line:introul-start}^ana opcon intro_unlabeled {
-  	(* translate intro operator to nested pairs *)
-  	fn (ty :: Ty, param :: unit, args :: list(Arg)) => tycase(ty) {
-        c(param) => (fold (zipExact (list_of_ltom param) args)
-      	  `SCSS()ECSS`
-          (fn (((lbl, ty), arg), cur_trans) => `SCSS(%{ECSSana(arg, ty)SCSS}, %ECSScur_transSCSS)ECSS`)
-        )
-      | _ => raise TyErr("...")
-    }
-  }^\label{line:introul-end}^
-
-  ^\label{line:introl-start}^syn opcon intro_labeled {
-  	fn (param :: list(Lbl), args :: list(Arg)) => 
-      (* ... same as intro_unlabeled but reorder first ... *)
-  }^\label{line:introl-end}^
-
-  ^\label{line:prj-start}^syn opcon # {
-  	fn (param :: Lbl, args :: list(Arg)) =>  
-      (* ... generate the appropriate n-fold projection ... *)
-  }^\label{line:prj-end}^
-
-  ^\label{line:ltpl-conc-start}^syn opcon + {
-  	fn (param :: unit, args :: list(Arg)) => 
-      (* generate the new type and combine the two nested tuples *)
-  }^\label{line:ltpl-conc-end}^
-
-  ^\label{line:ltpl-drop-start}^syn opcon - {fn (param : Lbl, args : list(Arg)) => 
-    (* generate the new type and drop the appropriate element *)
-  }^\label{line:ltpl-drop-end}^
-} with syntax ltplx for c ^\label{line:ltplx-annotation}^
-(* synonyms used in Sec. 5.1 *)
-^\label{line:tycon-synonym}^let tycon ltuplex = LTupleX.c
-let opcon # = LTupleX.#
-let opcon ltuple+ = LTupleX.+ 
-let opcon ltuple- = LTupleX.-
-\end{lstlisting}
-\caption{Using metamodules to define the type structure of labeled tuple types.}
-\label{fig:ltuplex}
-\end{figure}
-An example of a metasignature, \lstinline{LTUPLEX}, and a matching 
-metamodule, \lstinline{Ltuplex}, is shown in Figure \ref{fig:ltuplex}.
-We also show a \emph{metamodule-parameterized TSM} \lstinline{ltplx} to decrease syntactic cost.
-The remainder of this subsection explains these definitions.
-
-\paragraph{Type Constructors} On line \ref{line:ltuplex-c-sig}, the metasignature \lstinline{LTUPLEX} specifies a type constructor \lstinline{c} parameterized by static values of kind \lstinline{LblTyOrdMap}.
-This user-defined kind (not shown here for brevity) classifies ordered mappings from labels (of kind \lstinline{Lbl}, also user-defined but not shown) to types, which are static values of kind \lstinline{Ty} (a primitive kind).
-Let us assume that we have defined a \emph{kind-specific language} for \lstinline{LblTyOrdMap}, so we can write a static value of this kind concretely like this:
-\begin{lstlisting}[numbers=none]
-let static val Paper_ty_idx :: LblTyOrdMap = {
-  title : rstring /.+/
-   conf : rstring /([A-Z]+) (\d\d\d\d)/
-}
-\end{lstlisting}
-
-We can construct a type by applying the type constructor \lstinline{Ltuplex.c} to this parameter:
-\begin{lstlisting}[numbers=none]
-let type Paper = Ltuplex.c Paper_ty_idx
-\end{lstlisting}
-Line \ref{line:tycon-synonym} defines the tycon synonym \lstinline{ltuplex} for \lstinline{Ltuplex.c}.
-If we also substitute in the value of \lstinline{Paper_ty_idx} just given, we see that we have recovered the definition of \lstinline{Paper} given at the beginning of Sec.
+
+Figure \ref{fig:ltuplex} defines a metasignature, \lstinline{LTUPLE}, and a matching 
+metamodule, \lstinline{Ltuple}.
+It also defines a \emph{metamodule-parameterized TSM} \lstinline{ltpl}, which will give us the introductory syntax shown in Sec.
+\ref{sec:metamodules-example}.
+The remainder of this subsection explains this figure.
+
+\paragraph{Type Constructors} On line \ref{line:ltuplex-c-sig} of Figure \ref{fig:ltuplex}, the metasignature \lstinline{LTUPLE} specifies that all matching metamodules must define a type constructor \lstinline{c} parameterized by {static values} of {sort} \lstinline{ComponentSpec}.
+%This user-defined sort (not shown) classifies ordered mappings from labels (of sort \lstinline{Lbl}, also user-defined but not shown) to types, which are static values of sort \lstinline{Ty} (a primitive sort).
+Let us assume that we can write static values of sort \lstinline{ComponentSpec} concretely like this (in the dissertation, we will discuss how it would be possible to adapt TSLs to also operate at the level of static values):
+\begin{lstlisting}[numbers=none]
+let static val Paper_tyidx :: ComponentSpec = {
+  SPCTtitle : EPCTrstring /SURL.+EURL/SPCT,EPCT
+   SPCTconf : EPCTrstring /SURL([A-Z]+) (\d\d\d\d)EURL/
+}
+\end{lstlisting}
+Applying the type constructor \lstinline{Ltuple.c} to this parameter gives us a type:
+\begin{lstlisting}[numbers=none]
+let type Paper = Ltuple.c Paper_tyidx
+\end{lstlisting}
+Line \ref{line:tycon-synonym} defines the tycon synonym \lstinline{ltuple} for \lstinline{Ltuple.c}.
+Substituting into the above, we recover the definition of \lstinline{Paper} given at the beginning of Sec.
 \ref{sec:metamodules-example}:
 \begin{lstlisting}[numbers=none]
-let type Paper = ltuplex {
-  title : rstring /.+/
-   conf : rstring /([A-Z]+) (\d\d\d\d)/
-}
-\end{lstlisting}
-Note that because types are static values, this is equivalent to writing:
-\begin{lstlisting}[numbers=none]
-let static val Paper :: Ty = ltuplex {
-  title : rstring /.+/
-   conf : rstring /([A-Z]+) (\d\d\d\d)/
-}
-\end{lstlisting}
-
-Recall from Sec.
-\ref{sec:verse} that each valid external type has a translation, which is an internal type.
-The metamodule \lstinline{Ltuplex} defines a \emph{type translation schema}, a static function that determines the translations of types constructed by \lstinline{c} (each identified by the value of its parameter).
-Internal types are encoded as static values of kind \lstinline{ITy}, which we assume supports standard \emph{quasiquotation} syntax.
-Here, the  type translation schema on lines \ref{line:c-start} to \ref{line:c-end} generates nested internal products by folding over a list generated from the type parameter (in the standard way), returning the encoding of \lstinline{unit} for the empty case, and recursively nesting encodings of binary products otherwise using the standard \emph{unquote} operator, \lstinline{%}.
-For example, the encoding of the type translation of \lstinline{Paper} determined by this static function is:
+let type Paper = ltuple {
+  SPCTtitle : EPCTrstring /SURL.+EURL/SPCT,
+   conf : EPCTrstring /SURL([A-Z]+) (\d\d\d\d)EURL/
+}
+\end{lstlisting}
+Note that because types are static values of sort \lstinline{Ty}, this is equivalent to writing:
+\begin{lstlisting}[numbers=none]
+let static val Paper :: Ty = ltuple {
+  SPCTtitle : EPCTrstring /SURL.+EURL/SPCT,
+   conf : EPCTrstring /SURL([A-Z]+) (\d\d\d\d)EURL/
+}
+\end{lstlisting}
+
+Also note that to ensure that type equality is decidable, tycon parameters can only be of a sort for which equality coincides with structural comparison of values.
+We call these sorts \emph{equality sorts}.
+Equality sorts are exactly analagous to equality types as found in Standard ML.
+The main implication of this restriction is that type parameters cannot contain static functions (because they cannot be compared structurally).
+
+\paragraph{Type Translations} Recall from Sec.
+\ref{sec:verse} that each external type must translate to an internal type.
+Each metamodule controls the translation of types constructed by its tycons by defining a static function called the \emph{type translator}.
+This static function programmatically generates the type's translation given its type parameter.
+Type translations are represented as static values of sort \lstinline{ITy}, which we assume supports standard \emph{quasiquotation} syntax.
+
+The  type translator for \lstinline{Ltuple.c} is defined on lines \ref{line:c-start}-\ref{line:c-end} of Figure \ref{fig:ltuplex}.
+We have chosen here to translate labeled tuple types to nested binary product types internally.
+The type translator generates these by folding over the component mapping (using helper function \lstinline{list_of_comp_spec :: ComponentSpec -> list(Lbl * Ty)}, not shown).
+It generates the representation of \lstinline{unit} for the empty case, and nests the binary product types in the recursive case (using the standard \emph{unquote} syntax, \lstinline{%}).
+For example, the representation of the type translation of \lstinline{Paper} determined by this static function is:
 \begin{lstlisting}[numbers=none]
@@
 Notice that the translations of the component regular string types are not inserted directly, but are rather treated {parametrically} using the \emph{translational form} \lstinline{SCSStrans(ECSSTSCSS)}.
-This will be key to the modularity principle that we plan to explore in the remaining work.
-
-If, for example, all regular string types translate to standard internal strings, encoded by some internal type abbreviated \lstinline{string}, and substitute this in, then the translation of \lstinline{Paper} is ultimately:
+This will be key to the modularity principle -- \emph{translation independence} -- that we will discuss shortly.%\todo{discuss SL}
+
+Let us assume, for example, that all regular string types translate to standard internal strings, encoded by some internal type abbreviated \lstinline{string}.
+After we substitute this in for these translational forms, we arrive at the translation of \lstinline{Paper}:
 \begin{lstlisting}[numbers=none]
@@
 
-\paragraph{Operator Constructors} On line \ref{line:intro-unlabeled-sig} of Figure \ref{fig:ltuplex}, the metasignature \lstinline{LTUPLEX} specifies an operator constructor, \lstinline{intro_unlabeled}, parameterized by static values of kind \lstinline{unit}.
-The qualifier \lstinline{ana} specifies that this opcon can only be used in analytic positions (i.e.
+Formally, we should be able to derive that $\vdash \texttt{Paper}~\mathtt{type} \leadsto \texttt{string} \times (\texttt{string} \times \mathsf{unit})$. The following rule governs type translation:
+\begin{mathpar}
+\inferrule[trans-ty]{
+    \vdash_\Phi \mathbf{ty}(m\cdot c; \sigma_\text{param}) :: \mathsf{Ty}\\
+    \mathbf{ty}(m\cdot c; \sigma_\text{param})~\mathtt{sval}_\Phi\\
+	\mathbf{translator}(\Phi; m\cdot c)=\sigma_\text{translator}\\
+	\sigma_\text{translator}(\sigma_\text{param}) \Downarrow \sigma_\text{trans}\\
+	\sigma_\text{trans} \uparrow_\Phi^m \tau_\text{abstrans} \parallel {\delta} : {\Delta}\\
+	{\Delta} \vdash \tau_\text{abstrans}~\mathtt{itype}
+}{
+	\vdash_\Phi \mathbf{ty}(m.c;\sigma_\text{param})~\mathtt{type}\leadsto[{\delta}]\tau_\text{abstrans}
+}
+\end{mathpar}
+Note that this rule is again slightly simplified here.
+It would need to be modified to handle external type abstraction, for example.
+However, it is pedagogically useful to begin with this simpler rule.
+Note that our purpose in this proposal is not to give a complete formal account of our contributions, but only to sketch out what this formal account will look like.
+
+Looking at the conclusion of the rule, we see that in the abstract syntax of the SL, types constructed by user-defined tycons take the form $\textbf{ty}(m\cdot c; \sigma_\text{param})$, where $m\cdot c$ refers to a  {tycon} $c$ defined by metamodule $m$ and $\sigma_\text{param}$ is the type parameter.
+The premises can be understood as follows:
+\begin{enumerate}
+\item The first premise checks that the type is of the required sort, $\mathsf{Ty}$.
+This involves finding the parameter sort associated with the tycon in the \emph{metamodule context}, $\Phi$ (we omit the straightforward definitions and rules for concision here).
+\item The second premise checks that the parameter is a static value. %All static expressions, including types, are fully evaluated during typechecking and translation.
+\item The third premise looks up the translator associated with $m \cdot c$ in $\Phi$. 
+\item The fourth premise applies this translator to the type parameter to generate the representation of the type translation, $\sigma_\text{trans}$ (e.g. shown above for \lstinline{Paper}).
+\item The fifth premise decodes $\sigma_\text{trans}$ to produce an \emph{abstracted type translation}, $\tau_\text{abstrans}$.
+It is ``abstracted'' in that translational forms referring to types constructed by metamodules other than $m$, e.g.
+\lstinline{trans(rstring /.+/)} above, have been replaced by type variables.
+For example, the representation above produces the abstracted typed translation $\alpha_1 \times (\alpha_2 \times \mathsf{unit})$.
+The actual translations of these types are packaged into a standard \emph{type substitution}, $\delta$.
+For example, here $\delta = [\texttt{string}/\alpha_1, \texttt{string}/\alpha_2]$.
+The context corresponding to this substitution, here $\Delta=\alpha_1, \alpha_2$, is also generated.
+\item The final premise validates the abstracted type translation under $\Delta$ by making sure it is a valid internal type according to the internal statics.
+\item Finally, the conclusion of the rule applies $\delta$ to the abstracted translation to produce the final type translation.
+\end{enumerate}
+
+We will see why having this ability to selectively hold type translations abstract is critical to modularity after we consider operator constructors.
+
+\paragraph{Operator Constructors} On line \ref{line:intro-unlabeled-sig} of Figure \ref{fig:ltuplex}, the metasignature \lstinline{LTUPLE} specifies an operator constructor, \lstinline{intro_unlabeled}, parameterized by static values of sort \lstinline{unit}.
+The qualifier \lstinline{ana} specifies that this opcon is only for use in analytic positions (i.e.
 where the expected type is known).
-For example, \lstinline{Ltuplex.intro_unlabeled} can be applied like this, because the return type of the function determines the expected type:
+For example, we can apply \lstinline{Ltuple.intro_unlabeled} like this, because the return type of the function determines the expected type:
 \begin{lstlisting}[numbers=none]
 fun make_paper(conf : conf_t, title : title_t) : Paper => 
-  Ltuplex.intro_unlabeled () (conf; title)
-\end{lstlisting}
-First, a static parameter value of the appropriate kind is supplied. After that, an \emph{argument list}, \lstinline{(conf; title)}, is supplied.
-
-The metamodule \lstinline{Ltuplex} determines how this expression is typechecked and translated on lines \ref{line:introul-start}-\ref{line:introul-end}.
-Analytic opcons like \lstinline{intro_unlabeled} are defined with a static function that is given the type that the expression is being analyzed against, here \lstinline{Paper}, the operator parameter, here \lstinline{()}, and a list of \emph{argument interfaces}, which will allow it to ask the semantics to recursively typecheck and translate the provided arguments.
-The function must return an encoding of an  internal expression, which will become the translation of the external expression.
-Internal expressions are encoded as static values of kind \lstinline{IExp}, and again we assume that we can use standard quasiquote syntax.
-Here, we simply generate the trivial value for an empty labeled tuple and recursively generate nested pairs otherwise, following directly the structure for the type translation generator above.
-This is, of course, intentional: the introductory operation for a type must generate an expression translation consistent with the corresponding type translation for type safety to hold.
-
-\todo{describe ltplx?}
+  Ltuple.intro_unlabeled () (conf; title)
+\end{lstlisting}
+First, we provide the opcon with a static parameter value of the appropriate sort, here \lstinline{unit}, to create an operator.
+Then, we supply the operator with an \emph{argument list}, \lstinline{(conf; title)}.
+
+This is syntactically unwieldy.
+Luckily, we can define a parameterized TSM, \lstinline{ltpl}, that gives us a more conventional syntax.
+As shown on lines \ref{line:ltplx-tsm-start}-\ref{ltplx-tsm-end} of Figure \ref{fig:ltuplex}, \lstinline{ltpl} defines syntax at all types that are of the form \lstinline{L.c(param)}, where \lstinline{L} is a metamodule matching \lstinline{LTUPLE} and \lstinline{param} is a static value of sort \lstinline{ComponentSpec}.
+In other words, the TSM defines syntax for all labeled tuple types.
+We elide the parser implementation, but as the comment indicates, we can rewrite \lstinline{make_paper} like this:
+\begin{lstlisting}[numbers=none]
+fun make_paper_tsm(conf : conf_t, title : title_t) : Paper => 
+  ltpl Ltuple Paper_tyidx {confSCSS, ECSStitle}
+\end{lstlisting}
+
+To avoid having to explicitly name the TSM and supply the parameters, we designate \lstinline{ltpl} as \lstinline{Ltuple.c}'s TSL on line \ref{line:ltplx-annotation}.
+Now we can rewrite \lstinline{make_paper} in the standard way:
+\begin{lstlisting}[numbers=none]
+fun make_paper_tsl(conf : conf_t, title : title_t) : Paper => 
+  {confSCSS, ECSStitle}
+\end{lstlisting}
+
+Putting syntax aside, the metamodule defining the opcon, \lstinline{Ltuple}, programmatically determines how to typecheck and translate this operation, again by the defining a static function called the \emph{opcon definition}.
+The definition of \lstinline{intro_unlabeled} is shown on lines \ref{line:introul-start}-\ref{line:introul-end} of Figure \ref{fig:ltuplex}.
+For analytic opcons like \lstinline{intro_unlabeled}, the semantics provides the opcon definition  with the type that the expression is being analyzed against, here \lstinline{Paper}, the operator parameter, here \lstinline{()}, and a list of \emph{argument interfaces}, which will allow it to ask the semantics to recursively typecheck and translate the arguments provided in the argument list.
+It must return a representation of an  internal expression, which will become the translation of the operation.
+Internal expressions are represented as static values of sort \lstinline{IExp}, and again we assume that we can use standard quasiquote syntax.
+
+On lines \ref{line:introul-start}-\ref{line:introul-end}, the opcon definition first checks that the type is actually a labeled tuple type using the \lstinline{tycase} primitive, which case analyzes against the type constructor of the provided type.
+There must always be a default case to ensure exhaustiveness.
+In this example, the default case raises a type error (with an error message, elided here).\footnote{The reason why we use this exception-like data flow for raising type errors, rather than using an option sort, is somewhat subtle so we defer it for discussion in the dissertation.} If the tycon is \lstinline{c}, we pair up the items in the type's component specification with the arguments and fold over this list.
+The empty case generates the representation of the trivial internal value and the recursive case generates nested pairs.
+Notice that this follows the structure of the type translator for \lstinline{c}, reflecting the fact that introductory operations at a type must generate a translation consistent with the corresponding type translation for type safety to hold.
+The static operation \lstinline{ana(arg, ty)} requests that the provided argument be analyzed against the provided type, evaluating to a representation of its translation if this succeeds and raising an error if it fails.
+For our example above, the representation of the translation generated by the opcon definition is:
+\begin{lstlisting}[numbers=none]
+`SQT(anatrans[0](EQTconf_tSQT), (anatrans[1](EQTtitle_tSQT), ()))`
+\end{lstlisting}
+An \emph{argument translation} of the form \lstinline{SQTanatrans[i](EQTTSQT)} stands in for the translation of argument \lstinline{i} analyzed against type \lstinline{T}.
+
+The semantics next generates an \emph{abstracted translation}, which is a corresponding internal term where each argument translation has been replaced by a corresponding fresh variable.
+To validate it, we assume that the type of each variable is the abstracted type translation of the type the corresponding argument was analyzed against.
+So in this case, the abstracted translation is $(x_0, (x_1, ()))$ and it will be validated under a context where $x_0 : \alpha_0$ and  $x_1 : \alpha_1$ against the abstracted type translation for the type the operation as a whole was being analyzed against, i.e.
+$\alpha_0 \times (\alpha_1 \times \mathsf{unit})$, as discussed above.
+Only after this succeeds are the actual argument translations substituted in to generate the final operation translation, here $({conf}, ({title}, ())$.
+
+By treating the arguments to the operation parametrically in this way, we can be sure that the metamodule defining labeled tuple types cannot violate the translation invariants that other metamodules maintain.
+For example, even if regular strings are implemented (by some other metamodule, not shown) as strings satisfying the regular string invariant, \lstinline{Ltuple} knows only that \emph{there exists} some translation for each regular string type, but nothing more.
+If the regular string metamodule later decides to change how it implements regular string types, this would require only local changes to that metamodule.
+We call this property \emph{translation independence}, by analogy to the \emph{representation independence} property that underlies modular reasoning with ML modules.
+An important point is that we are able to achieve modularity guarantees without requiring mechanized proofs of translation invariants (just as ML does not require mechanized proofs of representation invariants).
+%Translation independence allows library providers to reason locally about \emph{translation invariants} maintained by the operators in a metamodule.
+For example, a metamodule defining regular string types would maintain the regular string invariant.
+No other metamodule, by virtue of this translation independence property, can violate this invariant.
+
+To formalize the intuition just described, we would need to introduce a number of technical devices.
+It would require too much space to give a comprehensive formal account of this mechanism in this proposal, so  we leave this as work that remains to be done (though see below for progress).
+In broad strokes, it is similar to the rule for type translation described above.
+
+% \begin{mathpar}
+% \inferrule[ap-anaop]{
+%   \mathbf{ana opcon}~m.\text{op}~\mathbf{of}~\uppi_\text{param}~\{ \sigma_\text{opdef} \} \in \Phi\\
+%   \vdash_\Phi \sigma_\text{param} :: \uppi_\text{param}\\
+%   \mathbf{args}(\es) = \sigma_\text{args}\\
+%   \sigma_\text{opdef} (\sigma_\text{ty}) (\sigma_\text{param}) (\sigma_\text{args}) \Downarrow_{\hat{\Gamma};\Phi;\es} \sigma_\text{trans}\\
+%   \sigma_\text{trans} \parallel \emptyset~\emptyset \uparrow_{\hat{\Gamma}; \Phi; \es}^m \tau_\text{abstrans} \parallel \memD~\memG\\
+%   \mathsf{trans}(\sigma_\text{ty}) \parallel \memD \uparrow_{\Phi}^m 
+% }{
+% 	\hat{\Gamma} \vdash_\Phi \mathbf{opconap}(m; \text{op}; \sigma_\text{param}; \bar{e}) \Leftarrow \sigma_\text{ty} \leadsto [\delta][\gamma]\iota
+% }
+% \end{mathpar}
+% omega anaop of pi_param { sigma_opdef }
+% sigma_param :: pi_param
+% arginterface(es) = sigma_args
+% sigma_opdef (sigma_ty) (sigma_param) (sigma_args) |v sigma_trans
+% sigma_trans || 0 0 ^| iota_trans |\ (g : G) (d : D)
+% trans(sigma) || d : D ^| tau_tytrans || d' : D'
+% D G |- iota_trans : tau_tytrans
+% -----------------------------------------------------------------
+% opconap(omega; sigma_param; es) <= sigma ~> [delta][gamma]iota
 
 The remaining opcons are synthetic opcons, which means they can be used anywhere.
-They operate similarly, but rather than taking a type as input, they produce a pair of a type and a translation.
-For concision, we omit the details.
-Instead, note that the code written in these definitions is essentially exactly what would appear in a standard implementation of the first stage of a compiler.
-The novelty is not in the definitions, but in how they are organized and brought into the language.
-\subsection{Timeline}\label{sec:metamodules-timeline}
-For the sake of concision, we leave providing substantially more technical details on how metamodules operate as work that remains to be done.
-However, we have a lengthy draft of a technical report that provides these details (for an older version of this work).
-We anticipate completing this work will take about 2 months during Summer 2015.
-
+They operate similarly, but rather than taking a type as input, they produce a pair of a type and a translation as output.
+In every other way, these are analagous to analytic opcons, so we omit the details here.
+Instead, note that the static functions that define these opcons contain code that looks just like the code that would appear in a standard implementation of the first stage of a compiler for a language that built in labeled tuple types.
+The novelty is not in these definitions, but in how they are organized and brought into the language.
+
+\subsubsection{Limitations}
+The main limitation of the mechanism just described is that it only supports defining types and operators that do not require adding new contexts to the typing judgement.
+No new binding structure can be defined.
+Fortunately, a large number of interesting examples have this character.
+We plan to discuss this as an avenue for future work in the thesis.
+
+The metatheory is also somewhat tricky, again because it is not an immediate application of rule induction.
+We plan to discuss an induction principle by which we can prove the necessary metatheorems in the dissertation.
+
+\subsection{Timeline and Milestones}\label{sec:metamodules-timeline}
+The work above has recently been refactored so as to more closely follow the construction of the ML module system.
+Previously, the main organizational unit was a single tycon, rather than a metamodule.
+In this alternative form, we have a complete paper draft and a nearly complete technical report that gives the details of all of the mechanisms described above.
+They can be found at:
+\begin{itemize}
+\item Paper:\\ \url{https://github.com/cyrus-/papers/blob/master/atlam-icfp15/atlam-icfp15.pdf}
+\item Technical Report:\\ \url{https://github.com/cyrus-/papers/blob/master/atlam-icfp15/atlam-icfp15-supplement.pdf}
+\end{itemize}
+
+We propose the following concrete milestones, to be completed after the work in the previous section:
+\begin{enumerate}
+\item Refactor the syntax around metamodules \textbf{(1 week)}
+\item Update the static language's semantics \textbf{(1 week)}
+\item Update the external language's semantics \textbf{(1 week)}
+\item Update and complete the metatheory \textbf{(3 weeks)}
+\item Add support for type abstraction \textbf{(1 week)}
+\end{enumerate}
+
+To further validate our work, we plan to detail a number of other examples, including regular strings, case types, and a simple prototypic object system.
+We anticipate that writing these up, along with other writing tasks, will take another \textbf{2 weeks}.
+So the total for this section is 9 weeks.
 % The semantics uses internal type abstraction to guarantee that metamodule providers can reason modularly about translation invariants, just as it uses external type abstraction to guarantee that module providers can reason modularly about representation invariants.%For example, a metamodule that expresses the type structure of records wouldLblTyList introduce a type constructor  \texttt{record} parameterized by a static mapping from labels to types, and two operators: the introductory operator \texttt{intro}, parameterized by a list of row labels, and the row projection operator \texttt{prj}, parameterized by a row label.
@@
 By using %This simplifies the Verse core language relative to the core languages of ML,  Scala and other comparable languages.
+
+\section{Conclusion}
+In summary, we have proposed a new functional programming language, Verse, that gives programmers the ability to programmatically implement new derived concrete syntax and type structure using statically evaluated functions.
+Consequently, Verse does not need to build in constructs that comparable languages need to (or would need to) build in, like regular expression syntax, list syntax, labeled tuples and regular strings.
+These and other examples serve to validate our claim that Verse is highly expressive.
+
+We have also outlined a formal semantics for these novel primitives.
+A more detailed specification and metatheoretic results that validate our claims that Verse has a hygienic type discipline and strong modular reasoning principles represent the main avenues for remaining work.
+We have already made substantial progress on closely related variants of the mechanisms proposed here, so we anticipate that the work can be completed over the course of 17 weeks, per the timelines in the sections above.
+Given that the proposal should be complete by the end of Summer 2015, this implies that the remaining work can be completed over the course of Fall 2015.
+Final updates and the presentation of the dissertation should then be in early Spring 2016.
+
+I do not plan to write any papers until the committee understands my work well enough that our discussions are technical, rather than motivational, in nature.
 
@@
 must be expressive enough to allow users to associate rich run-time, compile-time and edit-time behaviors with user constructs directly, while being sufficiently restrictive to maintain the global safety properties of the language and system as a whole, and to ensure that constructs cannot interfere with one another.
+
+\begin{figure}[p!]
+\vspace{-25px}
+\begin{lstlisting}
+metasignature LTUPLE = metasig {
+  ^\label{line:ltuplex-c-sig}^tycon c of ComponentSpec
+  ^\label{line:intro-unlabeled-sig}^ana opcon intro_unlabeled of unit
+  ^\label{line:intro-labeled-sig}^syn opcon intro_labeled of list(Lbl)
+  ^\label{line:prj-sig}^syn opcon # of Lbl
+  ^\label{line:conc-sig}^syn opcon + of unit
+  ^\label{line:drop-sig}^syn opcon - of Lbl
+}
+
+^\label{line:ltplx-tsm-start}^syntax ltpl(L :: LTUPLE, param :: ComponentSpec) at L.c(param) {
+  static fn ps => (* 
+    - elaborates {e_1, ..., e_n} to 
+        L.intro_unlabeled () (e_1; ...; e_n)
+    - elaborates {lbl1 => e_1, ..., lbln => e_n} to 
+        L.intro_labeled [lbl1, ..., lbln] e_1 ... e_n
+  *)
+}^\label{ltplx-tsm-end}^
+
+metamodule Ltuple :: LTUPLE = metamod {
+  ^\label{line:c-start}^(* translate labeled tuple types to nested products *)
+  tycon c {
+  	(* type translation generator: *)
+  	static fn (param :: ComponentSpec) => fold (list_of_comp_spec param)
+      `SCSSunitECSS`
+      (fn ((lbl, ty), cur_ty_trans) => `SCSStrans(ECSStySCSS) * %ECSScur_ty_trans`)
+  }^\label{line:c-end}^
+
+  ^\label{line:introul-start}^ana opcon intro_unlabeled {
+  	(* translate intro operator to nested pairs *)
+  	static fn (ty :: Ty, op_param :: unit, args :: list(Arg)) => tycase(ty) {
+        c(ty_param) => (fold (zipExact (list_of_comp_spec ty_param) args)
+      	  `SCSS()ECSS`
+          (fn (((lbl, ty), arg), cur_trans) => `SCSS(%{ECSSana(arg, ty)SCSS}, %ECSScur_transSCSS)ECSS`)
+        )
+      | _ => raise TyErr("...")
+    }
+  }^\label{line:introul-end}^
+
+  ^\label{line:introl-start}^syn opcon intro_labeled {
+  	static fn (op_param :: list(Lbl), args :: list(Arg)) => 
+      (* ... same as intro_unlabeled but reorder first ... *)
+  }^\label{line:introl-end}^
+
+  ^\label{line:prj-start}^syn opcon # {
+  	static fn (op_param :: Lbl, args :: list(Arg)) =>  
+      (* ... generate the appropriate n-fold projection ... *)
+  }^\label{line:prj-end}^
+
+  ^\label{line:ltpl-conc-start}^syn opcon + {
+  	static fn (op_param :: unit, args :: list(Arg)) => 
+      (* generate the new type and combine the two nested tuples *)
+  }^\label{line:ltpl-conc-end}^
+
+  ^\label{line:ltpl-drop-start}^syn opcon - {
+  	static fn (op_param : Lbl, args : list(Arg)) => 
+      (* generate the new type and drop the appropriate element *)
+  }^\label{line:ltpl-drop-end}^
+} with syntax ltpl for c ^\label{line:ltplx-annotation}^
+(* synonyms used in Sec. 5.1 *)
+^\label{line:tycon-synonym}^let tycon ltuple = Ltuple.c
+let opcon # = Ltuple.#
+let opcon ltuple+ = Ltuple.+ 
+let opcon ltuple- = Ltuple.-
+\end{lstlisting}
+\caption{Using metamodules to define the type structure of labeled tuple types.}
+\label{fig:ltuplex}
+\end{figure}

### 2015-06-04 thesis-proposal/problem.tex
integrated jonathan's updates

@@
 Instead, its users are able to modularly  express a wide variety of useful constructs in libraries, using the comparatively small set of primitives that the language builds in.
-Functional programming languages like Standard ML (SML), OCaml and Haskell descend from mathematically elegant typed lambda calculi, diverging mainly in that they consider a number of pragmatic issues alongside  fundamental metatheoretic issues in their design.
-For example, they build in record types, generalizing the nullary and binary product types more common in minimal calculi, because explicitly labeled components can be cognitively useful.
-Similarly, they build in derived concrete syntax for common constructs like lists.% concision and readability.
+Functional programming languages like Standard ML (SML), OCaml and Haskell descend from mathematically elegant typed lambda calculi, diverging mainly in that they consider various pragmatic issues alongside  fundamental metatheoretic issues in the design of their syntax and semantics.
+For example, they build in record types, generalizing the nullary and binary product types more common in minimal calculi, because explicitly labeled components can be cognitively useful to programmers.
+Similarly, they build in derived concrete syntax for common library constructs, like lists.% concision and readability.
 %Minimal typed lambda calculi play a central role in the study of programming languages, and serve as direct inspiration for  functional programming languages like ML.
@@
 %The community around this language should, ideally, then be able to direct its efforts almost exclusively toward the development of useful libraries, rather than further modifications to the language itself.
-Unfortunately, a stable language design that actually achieves this ideal has yet to emerge, as evidenced by the diverse array of \emph{dialects} that continue to proliferate around all major contemporary languages.
+Unfortunately, a stable language design that fully achieves this ideal has yet to emerge, as evidenced by the diverse array of \emph{dialects} that continue to proliferate around all major contemporary languages.
 % For example, let us consider the many dialects of ML.
@@
 %Tools for  constructing new ``domain-specific'' dialects also continue to proliferate. % The situation is similar across all major language lineages. 
-In fact, tools that assist with the construction of so-called  ``domain-specific'' language dialects are only becoming increasingly popular. 
+Indeed, tools that assist with the construction of so-called  ``domain-specific'' language dialects are only becoming increasingly popular. 
 {This calls for an investigation}: why is it that programmers and researchers are so often unable to satisfyingly express the constructs that they need in libraries, as modes of use of the ``general-purpose'' primitives already available in major languages today?
@@
 %Syntactic cost is often assessed qualitatively \cite{green1996usability}, though quantitative metrics can be defined.
-Many other types of data can similarly benefit from the availability of specialized derived syntax.
+Many other types of data could similarly benefit from the availability of specialized derived syntax.
 For example, we will consider regular expression patterns expressed using abstract data types (as a mode of use of the ML module system) in Sec.
 \ref{sec:syntax}.
-We will be able to express precisely the semantics that we seek, but the usual approaches, e.g.
+We will be able to express precisely the semantics that we seek, but existing approaches, e.g.
 using dynamic string parsing, for approximating the concrete syntax of patterns will leave something to be desired.
@@
 
-More advanced dialects introduce new type structure, going beyond what is possible by context-free elaboration to the existing language.
+More advanced dialects introduce new type structure, going beyond what is possible with derived syntax.
 Record types are themselves a simple example -- they cannot be expressed by context-free elaboration to a language with only nullary and binary product types.
@@
 Furr and Foster describe a dialect of OCaml that builds in the type structure of C \cite{Furr:2005:CTS:1065010.1065019}.
-Tools like proof assistants and logical frameworks can be used to specify and reason metatheoretically about dialects like these, and tools like compiler generators lower the cost of implementing them, again contributing to their continued proliferation.
+Tools like proof assistants and logical frameworks are used to specify and reason metatheoretically about dialects like these, and tools like compiler generators and language frameworks \cite{erdweg2013state} lower the cost of implementing them, again contributing to their continued proliferation.
 
@@
 For example, a recent revision of Ocaml added support for  ``generalized algebraic data types'' (GADTs), based on research on guarded recursive datatype constructors \cite{XiCheChe03}.
-Using GADTs, Ocaml was able to move some of the \emph{ad hoc} machinery for typechecking operations that use format strings, like \texttt{sprintf}, out of the language and into a library (note however that syntactic machinery remains primitively built in).
+Using GADTs, Ocaml was able to move some of the \emph{ad hoc} machinery for typechecking operations that use format strings, like \texttt{sprintf}, out of the language and into a library (however, syntactic machinery remains primitively built in).
 
@@
 \section{Proposed Contributions}
-Our broad aim in the work being proposed is to introduce a new functional programming language called Verse\footnote{We distinguish Verse from Wyvern, which is the language referred to in prior publications about some of the work being proposed here, because Wyvern is a group effort evolving independently in some important ways.} that takes more radical steps down the second path just described by giving library providers the ability to directly express new derived concrete syntax and type structure of a variety of designs in a safe and modularly composable manner atop a minimal type-theoretic core.
-The result is that Verse is smaller than comparable languages like ML and Scala, and it renders dialect formation less frequently necessary.%This eliminates the needs to build in fewer \emph{ad hoc} constructs and dialects are less frequently necessary.
+Our broad aim in the work being proposed is to introduce a new functional programming language called Verse\footnote{We distinguish Verse from Wyvern, which is the language referred to in prior publications about some of the work being proposed here, because Wyvern is a group effort evolving independently in some important ways.} that takes more radical steps down the second path just described by giving library providers the ability to directly express new derived concrete syntax as well as new type structure of a variety of designs in a safe and modularly composable manner atop a minimal type-theoretic core.
+The result is that 1) Verse is smaller than comparable languages like ML and Scala, and 2) dialect formation is less often necessary.%This eliminates the needs to build in fewer \emph{ad hoc} constructs and dialects are less frequently necessary.
 %thereby reducing the need for language dialects and revisions.
@@
 the relevant rules in the static semantics call for the evaluation of \emph{static functions} that generate static representations of expressions and types.
-These are written in a \emph{static language} (SL).
+Library providers write static functions in the Verse \emph{static language} (SL).
+%Maintaining a separation between the static (or ``compile-time'') phase and the dynamic (or ``run-time'') phase is an important facet of Verse's design.
 % static code generation.
@@
 A functional programming language can give library providers the ability to %meta\-pro\-gram\-matic\-ally 
-implement new derived concrete syntax and type structure atop a minimal type-theoretic core while maintaining a hygienic type discipline and modular reasoning principles.
+implement new derived concrete syntax and type structure atop a minimal type-theoretic internal language while maintaining a hygienic type discipline and modular reasoning principles.
 %These  primitives are  expressive enough to subsume the need for a variety of primitives that are, or would need to be, built in to comparable contemporary languages.
@@
 It may also be useful to explicitly acknowledge that library providers could leverage the primitives we introduce   to define constructs that are in rather poor taste.
-We  expect that in practice, Verse will come with a standard library defining a carefully curated collection of standard constructs, as well as guidelines for advanced users regarding when it would be sensible to use the mechanisms we introduce (following the example of languages that support operator overloading or typeclasses, which also have the potential for ``abuse'').
+We  expect that in practice, Verse will come with a standard library defining a carefully curated collection of standard constructs, as well as guidelines for advanced users regarding when it would be sensible to use the mechanisms we introduce (following the example of languages that support operator overloading or type classes \cite{Hall:1996:TCH:227699.227700}, which also have the potential for such ``abuse'').
 For most programmers, using Verse should not be substantially different from using a comparable language or one of its dialects.%The vast majority of programmers should not use the primitives that we introduce directly.
 
-Finally, Verse intentionally does not support full-spectrum dependent types because they blur the phase separation between ``compile-time'' and ``run-time''.
-Verse is designed for use in situations where SML, Ocaml, Haskell or Scala would be used today.
-That said, we conjecture that the primitives we describe could be added to languages like Gallina (the ``external language'' of the Coq proof assistant) with some modifications.
+Finally, Verse intentionally does not support dependent types ala Coq, Agda or Idris because they lack a  phase separation between ``compile-time'' and ``run-time.'' Verse is designed for use for programming tasks where SML, Ocaml, Haskell or Scala would be used today.
+That said, we conjecture that the primitives we describe could be added to languages like Gallina (the ``external language'' of the Coq proof assistant  \cite{Coq:manual}) or to the ``program extraction'' mechanisms of proof assistants like Coq with some modifications, but do not plan to pursue this line of research here.
 %Our interest is in making sure that the standard library is no more privileged than any other library.
@@
 \subsubsection{Limitations}
-The main limitation of the mechanism just described is that it only supports defining types and operators that do not require adding new contexts to the typing judgement.
+The main limitation of the mechanism just described is that it only supports defining types and operators that do not require adding new contexts to the EL typing judgement.
 No new binding structure can be defined.
-Fortunately, a large number of interesting examples have this character.
-We plan to discuss this as an avenue for future work in the thesis.
-
-The metatheory is also somewhat tricky, again because it is not an immediate application of rule induction.
-We plan to discuss an induction principle by which we can prove the necessary metatheorems in the dissertation.
+Fortunately, a number of interesting examples found in existing languages have this character, including labeled tuple types, record types, labeled sum types, object types (of various designs), regular string types and range-restricted numeric types.
+We plan to detail these in the thesis.
+On the other hand, constructs like ML-style exceptions, Ocaml-style open datatypes or Java-style class-based object systems would not be possible because they require statically tracking which constructors/classes are in scope.
+We plan to discuss adding support for new typing contexts as an avenue for future work in the thesis.
+
+Establishing the metatheory of metamodules is also somewhat tricky, again because it involves reasoning about expressions that are not clearly subexpressions of those in the conclusion of a rule.
+We plan to give an induction principle by which we can prove the necessary metatheorems in the dissertation.
 
@@
 \item Update and complete the metatheory \textbf{(3 weeks)}
-\item Add support for type abstraction \textbf{(1 week)}
+\item Add support for external type abstraction \textbf{(1 week)}
 \end{enumerate}
 
-To further validate our work, we plan to detail a number of other examples, including regular strings, case types, and a simple prototypic object system.
-We anticipate that writing these up, along with other writing tasks, will take another \textbf{2 weeks}.
+To further validate our work, we plan to detail a number of the examples mentioned above, including regular strings, case types, and a simple prototypic object system.
+We anticipate that the writing involved to do this, along with other writing tasks, will take another \textbf{2 weeks}.
 So the total for this section is 9 weeks.
@@
 Consequently, Verse does not need to build in constructs that comparable languages need to (or would need to) build in, like regular expression syntax, list syntax, labeled tuples and regular strings.
-These and other examples serve to validate our claim that Verse is highly expressive.
+These and other examples that we will cover in the dissertation will serve to validate our claim that Verse is highly expressive (and thus should lead to a decreased need for dialect formation).
 
@@
 Final updates and the presentation of the dissertation should then be in early Spring 2016.
-
-I do not plan to write any papers until the committee understands my work well enough that our discussions are technical, rather than motivational, in nature.
 

### 2015-06-10 thesis-proposal/problem.tex
addressed jonathan's comments thru sec 4

@@
 To decrease syntactic cost, library providers construct \emph{derived syntactic dialects}, i.e.
-dialects that are specified by context-free elaboration to the existing language.
+dialects that are specified by context-independent elaboration to the existing language.
 For example, Ur/Web is a  derived syntactic dialect of Ur (a language that itself descends from ML) that builds in derived syntax (colloquially, ``syntactic sugar'') for SQL queries, HTML elements and other datatypes used for web programming \cite{conf/popl/Chlipala15}.
 %Syntactic cost is often assessed qualitatively \cite{green1996usability}, though quantitative metrics can be defined.
-Many other types of data could similarly benefit from the availability of specialized derived syntax.
-For example, we will consider regular expression patterns expressed using abstract data types (as a mode of use of the ML module system) in Sec.
+These are far from the only types of data that could similarly benefit from the availability of specialized derived syntax.
+As another example, we will consider regular expression patterns expressed using abstract data types (as a mode of use of the ML module system) in Sec.
 \ref{sec:syntax}.
-We will be able to express precisely the semantics that we seek, but existing approaches, e.g.
-using dynamic string parsing, for approximating the concrete syntax of patterns will leave something to be desired.
+We will be able to express precisely the semantics that we seek, but existing approaches for approximating the concrete syntax of patterns, e.g.
+using dynamic string parsing,  will leave something to be desired.
 % An extension of Moscow ML provided syntax for monads encoded using the module system, inspired by the  syntax built in to Haskell.\todo{asked for reference, if not found will say that "Harper reports..."} %In fact, nearly all major programming languages primitively build in derived syntax for constructs that can otherwise be expressed in the standard library, e.g.
@@
 More advanced dialects introduce new type structure, going beyond what is possible with derived syntax.
-Record types are themselves a simple example -- they cannot be expressed by context-free elaboration to a language with only nullary and binary product types.
-Various functional languages have explored ``record-like'' primitives that go still further, supporting functional update and extension operators, width and depth coercions (sometimes implicit)%\cite{Cardelli:1984:SMI:1096.1098}
-, methods, prototypic dispatch and other ``semantic embellishments'' that cannot be expressed by context-free elaboration to a language with only standard record types (we will detail an  example in Sec.
+Record types provide a simple example -- they cannot be expressed by context-independent elaboration to a language with only nullary and binary product types.
+Various  languages have explored ``record-like'' primitives that go still further, supporting functional update and extension operators, width and depth coercions (sometimes implicit)%\cite{Cardelli:1984:SMI:1096.1098}
+, methods, prototypic dispatch and other ``semantic embellishments'' that cannot be expressed by context-independent elaboration to a language with only standard record types (we will detail an  example in Sec.
 \ref{sec:metamodules}).
@@
 Furr and Foster describe a dialect of OCaml that builds in the type structure of C \cite{Furr:2005:CTS:1065010.1065019}.
-Tools like proof assistants and logical frameworks are used to specify and reason metatheoretically about dialects like these, and tools like compiler generators and language frameworks \cite{erdweg2013state} lower the cost of implementing them, again contributing to their continued proliferation.
+Tools like proof assistants and logical frameworks are used to specify and reason metatheoretically about dialects like these, and tools like compiler generators and language frameworks \cite{erdweg2013state} lower their implementation cost, again contributing to their proliferation.
 
@@
 \end{enumerate} 
-Both TSMs and metamodules are novel forms of \emph{static code generation}, i.e.
+Both TSMs and metamodules are novel forms of \emph{static code generation} (also often called \emph{static} or \emph{compile-time metaprogramming}), i.e.
 the relevant rules in the static semantics call for the evaluation of \emph{static functions} that generate static representations of expressions and types.
-Library providers write static functions in the Verse \emph{static language} (SL).
+Library providers write these static functions in the Verse \emph{static language} (SL).
 %Maintaining a separation between the static (or ``compile-time'') phase and the dynamic (or ``run-time'') phase is an important facet of Verse's design.
@@
 
-Finally, Verse intentionally does not support dependent types ala Coq, Agda or Idris because they lack a  phase separation between ``compile-time'' and ``run-time.'' Verse is designed for use for programming tasks where SML, Ocaml, Haskell or Scala would be used today.
-That said, we conjecture that the primitives we describe could be added to languages like Gallina (the ``external language'' of the Coq proof assistant  \cite{Coq:manual}) or to the ``program extraction'' mechanisms of proof assistants like Coq with some modifications, but do not plan to pursue this line of research here.
+Finally, Verse intentionally does not support dependent types ala Coq, Agda or Idris because they do not maintain a phase separation between ``compile-time'' and ``run-time.'' This phase separation is useful for programming tasks (where one would like to be able to discover errors before running a program, particularly ones that may have an effect) but less so for theorem provers (where it is mainly the fact that a pure expression is well-typed that is of interest).
+Verse is designed for use for programming tasks where SML, Ocaml, Haskell or Scala would be used today, not for sophisticated theorem proving.
+That said, we conjecture that the primitives we describe could be added to languages like Gallina (the ``external language'' of the Coq proof assistant  \cite{Coq:manual}) or to the ``program extraction'' mechanisms of proof assistants like Coq with  modifications, but do not plan to pursue this line of research in this dissertation.
 %Our interest is in making sure that the standard library is no more privileged than any other library.
@@
 \section{Verse}\label{sec:verse}
-To situate ourselves within a formal framework, let us begin with a brief overview of how Verse is organized and specified. Verse consists of a \emph{module language} atop a \emph{core language}.
+To situate ourselves within a formal framework, let us begin with a brief overview of how Verse is organized and specified. %Verse consists of a \emph{module language} atop a \emph{core language}.
 
@@
 The key novelty is that the core language is split into a user-facing \emph{typed external language} (EL) specified by type-directed translation to a minimal \emph{typed internal language} (IL).
-Notionally, one might compare the core language to the first stage of a type-directed compiler (e.g.
+The core language can notionally be compared to the first stage of a type-directed compiler (e.g.
 the TIL compiler for Standard ML \cite{tarditi+:til-OLD}) shifted ``one level up'' into the language itself.
@@
 This will  allow us to explicitly specify how Verse's \emph{local type inference} works, and in particular, how it interacts with the primitives that we aim to introduce.
-Like Scala, we intentionally do not seek to support global type inference.
+Like Scala, we intentionally do not seek to support non-local type inference.
 %, primarily because it is not technically compatible with the primitives we will introduce (but also because we believe that compilers for languages that support only local type inference generate clearer error messages).
@@
 Finally, the static language plays an essential role in the semantics of the novel primitives that we will introduce into the EL.
-The SL is another typed lambda calculus consisting of \emph{static expressions}, $\sigma$, classified by \emph{sorts}\footnote{We use this word to avoid confusion with the common (though redundant) phrase ``static type''.}, $\uppi$, according to the following judgements (the ``static statics''):
+The SL is another typed lambda calculus consisting of \emph{static expressions}, $\sigma$, classified by \emph{sorts}\footnote{We use this word to avoid confusion with the phrase ``static type''.}, $\uppi$, according to the following judgements (the ``static statics''):
 \\[1ex]
@@
 \sigma~\mathtt{sval} & \text{Static expression $\sigma$ is a static value.}\\
-\sigma~\mathtt{styerr} & \text{Static expression $\sigma$ raises a type error.}
+\sigma~\mathtt{styerr} & \text{Static expression $\sigma$ raises an external type error.}
 \end{array}
@@
 
-As suggested by the external type translation judgement above, external types are static values of a primitive sort written $\mathsf{Ty}$.
+As suggested by the form of the external type translation judgement above, external types are static values of a primitive sort written $\mathsf{Ty}$.
 We will return to how this works (and consider the metatheoretic implications of this choice) in Sec.
@@
 \subsection{Module Language}
+We will first specify our contributions assuming a language without an ML-style module system, then make modifications to this specification to make it compatible with the addition of a module system.
+
 The Verse module language is taken directly from Standard ML, with which we assume a working familiarity for the purposes of this proposal \cite{harper1997programming,MacQueen:1984:MSM:800055.802036} (a related module language, e.g.
@@
 We will give examples of its use in Sections \ref{sec:syntax} and \ref{sec:metamodules}, but because it has been thoroughly studied in the literature, we will defer to prior work both here and in the dissertation for most formal details.
-In particular, we will first specify our contributions assuming a language without an ML-style module system, then make modifications to this specification to make it compatible with the addition of a module system.
-To do so, we will assume that modules are being tracked by a module context, without detailing how this context is populated.
+In particular, W In doing so, we will assume that modules are being tracked by a module context, without detailing how this context is populated.
 %This context will be analagous to the \emph{named type context} specified in the papers just cited.% Instead, we will consider it abstractly where it interacts with the core language.%\todo{reminder that we are not going to re-specify the module system}
@@
 \section{Modularly Programmable Textual Syntax}\label{sec:syntax}
-To begin, let us assume a standard, ML-like semantics for the EL and consider its concrete syntax.
+To begin, let us assume a standard, ML-like semantics for the EL and SL and consider the EL's concrete syntax.
 Verse, like many major contemporary languages, specifies a textual concrete syntax for the EL.\footnote{Although Wyvern specified a layout-sensitive concrete syntax, to avoid unnecessary distractions, we will specify a more conventional layout-insensitive concrete syntax for Verse.} %We have chosen to specify a layout-sensitive textual concrete syntax (i.e.
@@
 \subsection{Motivating Example: Regular Expression Syntax}\label{sec:examples}
-Let us begin by taking the perspective of a regular expression library provider. We assume the reader has some familiarity with regular expressions \cite{Thompson:1968:PTR:363347.363387}.
+Let us begin by taking the perspective of a regular expression library provider.
+We assume the reader has some familiarity with regular expressions \cite{Thompson:1968:PTR:363347.363387}.
+We will discuss a standard variant of regular expressions that supports marking \emph{captured groups} (with parentheses in the concrete syntax) to make certain examples more interesting.
 
@@
   val case : (
-    'a -> 
-    (string -> 'a) ->
-    (t * t -> 'a) ->
-    (t * t -> 'a) ->
-    (t -> 'a) ->
-    (t -> 'a) -> 
+    t -> 
+    'a -> (* Empty case *)
+    (string -> 'a) -> (* Str case *)
+    (t * t -> 'a) -> (* Seq case *)
+    (t * t -> 'a) -> (* Or case *)
+    (t -> 'a) -> (* Star case *)
+    (t -> 'a) -> (* Group case *)
     'a)
@@
 The identity of this type is held abstract outside the module during typechecking (i.e.
-it acts as a newly generated type) so that the burden of proving that there is no way to use the case analysis function to distinguish patterns with the same normal form is local to the module.
+it acts as a newly generated type) so that the burden of proving that there is no way to use the case analysis function to distinguish patterns with the same normal form is local to the module, and implementation details do not escape (and can thus change freely).
 %The details are standard and not particularly relevant for our purposes, so we omit them here.
@@
 
-Some languages, anticipating such modes of use, build in alternative string forms that leave escape sequences uninterpretted.
+Some languages, anticipating such modes of use, build in alternative string forms that leave escape sequences uninterpreted.
 For example, Ocaml supports the following, which has only a constant syntactic cost:
@@
 \end{lstlisting}
-In other words, this approach leaves the well-formedness of patterns constructed from strings as a static verification condition. 
-Though violations of this condition can sometimes be discovered dynamically via testing, empirical data gathered from large open source projects suggests that there remain many malformed regular expression patterns that are not detected by a project's test suite ``in the wild'' \cite{spishak2012type}.
-
-Statically verifying that this condition holds throughout a program is quite difficult to automate in general, because it requires reasoning about arbitrary dynamic behavior.
-In this example, a decision procedure would need to be able to establish that the variable \lstinline{pattern} is equal to the parse function \lstinline{PU.parse}.
+Though ill-formed patterns can sometimes be discovered dynamically via testing, empirical data gathered from large open source projects suggests that there remain many ill-formed regular expression patterns that are not detected by a project's test suite ``in the wild'' \cite{spishak2012type}.
+
+Statically verifying that pattern formation errors will not dynamically arise is quite difficult to automate in general, because it requires reasoning about the full dynamics of the language.
+In this example, the verification procedure would need to be able to establish that the variable \lstinline{pattern} is equal to the parse function \lstinline{PU.parse}.
 If the string argument had not been written literally but rather computed, e.g.
 as \lstinline{"SSTR(GESTR" ^ "SSTRCESTR"} where \lstinline{^} is the string concatenation function applied in infix style, it would also need to be able to establish that this expression is equivalent to the string \lstinline{"SSTR(GCESTR"}.
-For patterns that are dynamically constructed based on input to a function (further discussed below), the problem is not solved by ``simply'' statically evaluating the expression.
-
-Of course, asking the client to provide a proof of well-formedness would completely defeat the purpose -- providing a concise concrete syntax.
+For patterns that are dynamically constructed based on input to a function (see below), the problem is not solved by ``simply''  evaluating the expression statically (or, more generally, in some earlier ``stage'' of evaluation).
+
+Of course, asking the client to provide a proof of well-formedness would completely defeat our purpose -- providing a concise concrete syntax.
 
@@
 \end{lstlisting}
-We needed to use both dynamic string parsing and explicit applications of pattern constructors to achieve the intended semantics.
+%We needed to use both dynamic string parsing and explicit applications of pattern constructors to achieve the intended semantics. 
 It is difficult to capture idioms like this more concisely using dynamic string parsing (we will see an example of syntax that does capture such idioms below).
@@
 Unfortunately, it introduces new problems.
-First, the systems mentioned thusfar cannot guarantee that \emph{syntactic conflicts} between such extensions will not arise.
+First, the systems mentioned thus far cannot guarantee that \emph{syntactic conflicts} between such extensions will not arise.
 As stated directly in the  Coq manual: ``mixing different symbolic notations in [the] same text may cause serious parsing ambiguity''.
@@
 However, note that the most natural starting tokens like \lstinline{pattern} cannot be guaranteed to be universally unique, so we would be forced to use a more verbose token like \lstinline{edu_cmu_verse_rx_pattern}.
-There is no principled way for clients of our extension to define local abbreviations for starting tokens because this mechanism is language-external.
+There is no way for clients of our extension to define scoped abbreviations for starting tokens because this mechanism operates purely on syntactic forms and has no notion of semantic scope.
 
@@
 
-For patterns that are constructed dynamically, we need to get a bit more creative.
+For patterns that are constructed dynamically, we need to get more creative.
 For example, we might repurpose the infix operators that are normally used for other purposes to support string and pattern splicing as follows:
@@
 
-While this does not leave us with syntax that is quite as distinct and concise as would be possible with a  na\"ive syntax extension,  it does avoid many of the problems with that approach: there cannot be syntactic conflicts (because the syntax is not extended at all), we can define macro abbreviations because macros are integrated into the language, there is a hygiene discipline that guarantees that the elaboration will not capture variables inadvertently, and using a typed macro system, there is a type discipline (i.e.
+Having to creatively repurpose existing syntax like this limits the improvements a library provider can make to syntactic cost (particularly when there are pre-existing conventions that are not consistent with the conventions adopted by the language).
+This also can create confusion for readers expecting parenthesized expressions to behave in a consistent manner.
+However,  this approach does avoid many of the problems of na\"ive syntax extensions: there cannot be syntactic conflicts (because the syntax is not extended at all), we can define macro abbreviations because macros are integrated into the language, there is a hygiene discipline that guarantees that the elaboration will not capture variables inadvertently, and using a typed macro system, there is a type discipline (i.e.
 programmers need not examine the elaboration to know what type a macro application must have).
@@
 TSMs defined at a single type, e.g.
-\lstinline{Pattern}, rather than a parameterized family of types):
+the case type \lstinline{Pattern}, rather than a parameterized family of types):
 \begin{mathpar}
 \inferrule[syn-aptsm]{
-    \Delta \vdash \psi ~@~ \sigma_\text{ty}~\{\sigma_\text{parser}\}\\
+    \vdash \psi ~@~ \sigma_\text{ty}~\{\sigma_\text{parser}\}\\
     \mathbf{parsestream}(b) = \sigma_\text{ps}\\
@@
     \sigma_\text{elab} \uparrow e_\text{elab}\\\\
-    \Delta; \Gamma; \emptyset; \emptyset \vdash e_\text{elab} \Leftarrow \sigma_\text{ty} \leadsto \iota_\text{trans}
+    \Gamma; \emptyset \vdash e_\text{elab} \Leftarrow \sigma_\text{ty} \leadsto \iota_\text{trans}
 }{
-    \Delta_\text{out}; \Gamma_\text{out}; \Delta; \Gamma \vdash \mathbf{aptsm}(\psi; b) \Rightarrow \sigma_\text{ty} \leadsto \iota_\text{trans}
+    \Gamma_\text{out}; \Gamma \vdash \mathbf{aptsm}(\psi; b) \Rightarrow \sigma_\text{ty} \leadsto \iota_\text{trans}
 }
@@
 Note also that we omit various contexts relevant only to other Verse primitives.
-We include only \emph{typing contexts}, $\Gamma$, which map variables to types, and \emph{type abstraction contexts}, $\Delta$, which track universally quantified type variables, both in the standard way.
-We describe why there are two of each -- the \emph{outer contexts} and the \emph{current contexts} -- below.
+We include only \emph{typing contexts}, $\Gamma$, which map variables to types in the standard way.
+We describe why there are two contexts -- the \emph{outer context} and the \emph{current context} -- below.
 
@@
 \begin{enumerate}
-\item The first premise can be pronounced ``TSM expression $\psi$ defines a valid TSM at type $\sigma_\text{ty}$ under $\Delta$ with parse function $\sigma_\text{parser}$''.
+\item The first premise can be pronounced ``TSM expression $\psi$ defines a valid TSM at type $\sigma_\text{ty}$ with parse function $\sigma_\text{parser}$''.
 The ``outputs'' of this judgement are the type $\sigma_\text{ty}$ (of sort $\mathsf{Ty}$) and the static parse function, $\sigma_\text{parser}$ (of sort $\mathsf{ParseStream} \rightarrow \mathsf{Exp}$), defined by the TSM.% We must omit the rules governing this judgement here (a complete specification of this judgement represents the main piece of remaining work, see below).
@@
 \item The second premise creates a parse stream, $\sigma_\text{ps}$, of sort $\mathsf{ParseStream}$, from the body of the delimited form.
-\item The third premise applies the parse function to the parse stream to generate an encoding of the elaboration, $\sigma_\text{elab}$, of sort $\mathsf{Exp}$.% Note that this is ``metaprogramming'' -- a static expression in the external semantics to generate an encoding of another external term.
+\item The third premise applies the parse function to the parse stream to generate a static representation of the elaboration, $\sigma_\text{elab}$ (of sort $\mathsf{Exp}$).% Note that this is ``metaprogramming'' -- a static expression in the external semantics to generate an encoding of another external term.
 \item The fourth premise decodes $\sigma_\text{elab}$, producing the elaboration itself, $e_\text{elab}$.
 \item The final premise \emph{validates} the elaboration by analyzing it against the type $\sigma_\text{ty}$ specified by the TSM.
-The {current} typing and type abstraction contexts are ``saved'' for use when a spliced subexpression is encountered during this process by setting them as the new \emph{outer contexts}, as indicated.
-
-Variables in the outer contexts are not directly available to the elaboration, e.g. they are ignored by the (syn-var) rule:
+The {current} typing context is ``saved'' for use when a spliced subexpression is encountered during this process by setting it as the new \emph{outer context}, as indicated.
+
+Variables in the outer context are not directly available to the elaboration, e.g. they are ignored by the (syn-var) rule:
 \begin{mathpar}
 \inferrule[syn-var]{ }{
-    \Delta_\text{out}; \Gamma_\text{out}; \Delta; \Gamma, x : \sigma \vdash x \Rightarrow \sigma \leadsto x
+    \Gamma_\text{out}; \Gamma, x : \sigma \vdash x \Rightarrow \sigma \leadsto x
 }
@@
 
-Outer contexts are switched back in when encountering marked spliced expressions (which arise as explained previously from spliced expressions in the parse stream):
+Outer contexts are switched back in when encountering marked spliced expressions:
 \begin{mathpar}
 \inferrule[syn-spliced]{
-    \emptyset; \emptyset; \Delta_\text{out}; \Gamma_\text{out} \vdash e \Rightarrow \sigma \leadsto \iota
+    \emptyset; \Gamma_\text{out} \vdash e \Rightarrow \sigma \leadsto \iota
 }{
-    \Delta_\text{out}; \Gamma_\text{out}; \Delta; \Gamma \vdash \mathbf{spliced}(e) \Rightarrow \sigma \leadsto \iota
+    \Gamma_\text{out}; \Gamma \vdash \mathbf{spliced}(e) \Rightarrow \sigma \leadsto \iota
 }
@@
 \inferrule[ana-spliced]{
-    \emptyset; \emptyset; \Delta_\text{out}; \Gamma_\text{out} \vdash e \Leftarrow \sigma \leadsto \iota
+    \emptyset; \Gamma_\text{out} \vdash e \Leftarrow \sigma \leadsto \iota
 }{
-    \Delta_\text{out}; \Gamma_\text{out}; \Delta; \Gamma \vdash \mathbf{spliced}(e) \Leftarrow \sigma \leadsto \iota
+    \Gamma_\text{out}; \Gamma \vdash \mathbf{spliced}(e) \Leftarrow \sigma \leadsto \iota
 }
@@
   (* ... *)
-} with syntax pattern at t
-\end{lstlisting}
-
- Local type inference then determines which TSM is implicitly applied when analyzing a delimited form not prefixed by a TSM name.
+} with syntax pattern at t (* technically, the 'at t' clause is redundant *)
+\end{lstlisting}
+
+\todo{clarify} Local type inference then determines which TSM is implicitly applied when analyzing a delimited form not prefixed by a TSM name.
 For example, this function is equivalent to \lstinline{example_tsm} (note the return type annotation):
@@
 Unlike labeled tuple types, record types are identified up to component reordering, so this mapping does not preserve certain type disequalities between labeled tuple types.
-It is not  possible to allow clients to omit component labels and rely on the positional information available in the type when introducing a value of record type for this reason.%, because this information is not maintained by the type system.
+It is not  possible to allow clients to omit component labels and rely on the positional information available in the type when introducing a value of record type for this reason.\todo{clarify}%, because this information is not maintained by the type system.
 

### 2015-06-12 thesis-proposal/problem.tex
addressed jonathan's comments thru sec 5

@@
 as \lstinline{"SSTR(GESTR" ^ "SSTRCESTR"} where \lstinline{^} is the string concatenation function applied in infix style, it would also need to be able to establish that this expression is equivalent to the string \lstinline{"SSTR(GCESTR"}.
-For patterns that are dynamically constructed based on input to a function (see below), the problem is not solved by ``simply''  evaluating the expression statically (or, more generally, in some earlier ``stage'' of evaluation).
+For patterns that are dynamically constructed based on input to a function (see below), the problem is not solved by ``simply''  evaluating the expression statically (or, more generally, in some earlier ``stage'' of evaluation) \cite{Jones:Gomard:Sestoft:93:PartialEvaluation}.
 
@@
 \subsection{Contributions}\label{sec:syntax-contributions}
-Verse provides a new external primitive -- the \textbf{typed syntax macro} (TSM) -- that combines the syntactic flexibility of syntax extensions with the reasoning guarantees of statically-typed macros.
-TSMs can be parameterized by modules, so they can be used to define syntax valid at all modules satisfying a specified signature.
+Verse provides a new external primitive -- the \textbf{typed syntax macro} (TSM) -- that combines the syntactic flexibility of syntax extensions with the reasoning guarantees of typed macros.
+TSMs can be parameterized by modules, so they can be used to define syntax valid at all types defined in modules satisfying a specified signature.
 As we will discuss in in Sec.
@@
 To further lower the syntactic cost of using TSMs, Verse also  supports \emph{type-specific languages} (TSLs).
-This allows library providers to associate a TSM directly with a generated type or parameterized type constructor.
+This allows library providers to associate a TSM directly with a generated type (or parameterized type constructor, more generally).
 For example, a module \lstinline{P} can associate \lstinline{pattern} with \lstinline{P.t} as follows:
@@
   (* ... *)
-} with syntax pattern at t (* technically, the 'at t' clause is redundant *)
-\end{lstlisting}
-
-\todo{clarify} Local type inference then determines which TSM is implicitly applied when analyzing a delimited form not prefixed by a TSM name.
-For example, this function is equivalent to \lstinline{example_tsm} (note the return type annotation):
+} with syntax pattern
+\end{lstlisting}
+
+When the semantics encounters a delimited form not prefixed by a TSM name, the TSM associated with the type it is being analyzed against is implicitly applied.
+For example, the following function is equivalent to \lstinline{example_tsm} (note the return type annotation):
 \begin{lstlisting}[numbers=none]
@@
 \end{lstlisting}
-From the client's perspective, it is essentially as if the language had built in derived syntax for lists and regular expression patterns directly.
-However, we did not need to build in this syntax primitively.%The only constraint is that this syntax must be used in an analytic position, which we argue is actually better for code compren when encountering unfamiliar syntax.
+From the client's perspective, it is essentially as if the language had built in derived syntax for lists and regular expression patterns directly.%However, we did not need to build in this syntax primitively.%The only constraint is that this syntax must be used in an analytic position, which we argue is actually better for code compren when encountering unfamiliar syntax.
 
@@
 %\todo{revise this section}The main limitation of this approach is that it 
-Elaboration validation ensures that an incorrect parser cannot threaten type safety, but proving type safety inductively is subtle because the elaboration is not a subexpression of the expression it will replace.
-There is a ``simple'' solution to this problem: the elaboration cannot itself use TSMs or TSLs.
+Though elaboration validation ensures that an incorrect parser cannot threaten type safety,  proving type safety inductively is subtle because the elaboration is not a subexpression of the expression it will replace.
+There is a ``simple'' solution to this problem: we impose the restriction that the elaboration cannot itself use TSMs or TSLs.
 This would be somewhat awkward in practice, so we plan to briefly discuss other more flexible solutions as well.
 
-Another concern is that if we define the SL such that it supports general recursion, this then causes typechecking to become undecidable, because the parse function may not terminate.
-In practice, this too should only be a minor concern -- a compiler might be expected to limit the running time of the parse function to ensure that compilation does not ``hang'' due to an incorrect parser.
+Another concern is that if we define the SL such that evaluating the parse function may not terminate, this then causes typechecking to become undecidable.
+In practice, this too should  be a minor concern.
+A compiler would be free to limit the running time of the parse function to ensure that compilation does not ``hang'' due to an incorrect parser.
 Related concerns would arise if the static language supported state or access to other external effects.
-Existing techniques for controlling effects would handle this problem, but this is not something we plan to discuss in detail here because our SL does not specify such effects.
+Existing techniques for controlling effects would handle this problem, but this is not something we plan to detail here because our SL does not specify these effects.
 
@@
 
-Finally, we have discussed expression syntax here, but have not explicitly considered pattern matching.
+Finally, we have discussed only expression syntax here, but have not considered pattern matching.
 In fact, we conjecture that it is completely straightforward to extend this work to allow a TSM to define both an expression parser and a pattern parser.
-We plan to concretely discuss this as the most immediate avenue for future work, but will not consider pattern matching in formal detail.
+We plan to informally discuss this as the most immediate avenue for future work, but will not consider pattern matching in formal detail in the dissertation.
 
 \subsection{Timeline and Milestones}\label{sec:syntax-timeline}
-We have described and given a more detailed formal specification of TSMs in a recently published paper \cite{sac15}, and TSLs in a paper published last year \cite{TSLs}, both in the context of the Wyvern language.
-The mechanics of statically invoking the parse function, elaboration encoding/decoding and hygienic elaboration validation have all been detailed there.
+We described and gave a more detailed formal specification of TSMs in a recently published paper \cite{sac15}, and TSLs in a paper published last year \cite{TSLs}, both in the context of the Wyvern language.
+The mechanics of defining a TSM, statically invoking the parse function, elaboration encoding/decoding and hygienic elaboration validation have all been detailed there.
 In the dissertation, I plan to first present the formal system closely following this presentation, updating it only to ensure that it is consistent with the work in the next section to present a unified design.
-The rule (syn-aptsm) above represents the most substantial step of this plan.
+The rule (syn-aptsm) above represents the first and most substantial step of this plan.
 I anticipate that this will take about \textbf{1.5 weeks} of further work.
@@
 Wyvern does not have an ML-style module system (type declarations are the only form of type generation) and at the time, it did not have parameterized types, so to actually support the examples we have given in this section, we also need to update the formalization to handle parameterized TSMs.
-We plan to flesh this out in the course of writing the dissertation, and I anticipate it requiring 5.5 weeks of my time after the proposal is complete.
-We propose the following concrete milestones:
+I plan to flesh this out in the course of writing the dissertation, and I anticipate it requiring 5.5 weeks of my time after the proposal is complete.
+I propose the following concrete milestones:
 
@@
 
-To validate our claims of expressiveness, we plan to give several more examples drawn from existing languages.
-We will include those we have worked out in the papers just cited as well as additional examples that highlight the benefits of integration with Verse's ML-based module system.
-For example, we will show how to implement Haskell's primitive ``do'' syntax using TSLs together with a modular encoding of monads (as described on Prof.
+To validate our claims of expressiveness, I plan to give several more examples drawn from existing languages.
+I will include those we have worked out in the papers just cited as well as additional examples that highlight the benefits of integration with Verse's ML-based module system.
+For example, I will discuss implementing Haskell's primitive ``do'' syntax using TSLs together with a modular encoding of monads (as described on Prof.
 Harper's blog \cite{SML/Monads}).
-We will also briefly summarize the corpus analysis we conducted in \cite{TSLs} where we found a large number of  situations in large publicly available Java libraries where dynamic string parsing was apparently being used to control syntactic cost.
-We anticipate that writing these examples up will take an additional \textbf{1 week} of work, bringing the total for this section up to 8 weeks.
+I will also show how parser generators and quasiquotation (e.g.
+as in Scala \cite{shabalin2013quasiquotes} or Lisp \cite{Bawd99a}) can be seen as modes of use of TSMs/TSLs.
+I will also briefly summarize the corpus analysis we conducted in \cite{TSLs} where we found a large number of  situations in large publicly available Java libraries where dynamic string parsing was apparently being used to control syntactic cost.
+I anticipate that writing these examples will take an additional \textbf{1 week} of work, bringing the total for this portion of the dissertation up to 8 weeks.
 %Adding in the time I will need to propose and some buffer time for unanticipated issues, my goal is to have this portion of the dissertation complete by the end of Summer 2015.
@@
 \end{lstlisting}
-The \lstinline{let type} construction defines a synonym, \lstinline{Paper}, for a type constructed by applying the \emph{type constructor} (or \emph{tycon}) \lstinline{ltuple} to a \emph{component specification}, which is a  statically valued ordered finite mapping from labels to types, written using conventional concrete syntax.
+The \lstinline{let type} construction defines a synonym, \lstinline{Paper}, for a type constructed by applying the \emph{type constructor} (or \emph{tycon}) \lstinline{ltuple} to a \emph{component specification}, which is a  statically valued (cf.
+Sec.
+2) ordered finite mapping from labels to types, written using conventional concrete syntax.
 
@@
 Regular string types are constructed by applying the tycon \lstinline{rstring} to a statically valued regular expression pattern, again written here using conventional concrete syntax like that discussed in the previous section.
-Regular string types classify expressions that behave like strings in the corresponding regular language, here the language of non-empty strings.
+Regular string types classify expressions that behave as strings in the corresponding regular language, here the language of non-empty strings.
 The second component of type \lstinline{Paper} is labeled \lstinline{conf} and specifies a regular string type with two {parenthesized groups}, corresponding to a conference abbreviation and year.
@@
 \end{lstlisting}
-Here, we've taken all possible valid introductory and projection operators and specified their semantics using function types.
-Expressing labels explicitly is an important facet of having labeled tuple types in a language, so we put them into the function identifiers.
-Even with our minimal EL, we could  implement the semantics of these operators against this signature using a Church-style encoding (the details are standard and omitted here).
-Alternatively, if we slightly relax  our insistence on minimalism and add binary products to the EL, we could use them to implement our desired semantics as well, like this:
+Here, we've taken all possible valid introductory and projection operators and called for their expression as functions.
+Expressing labels explicitly is an important facet of having labeled tuple types in a language, so we must put them into the function identifiers.
+Even with our minimal EL, we could  implement the semantics of the corresponding operators against this signature using a Church-style encoding (the details are standard and omitted here).
+Alternatively, if we slightly relax  our insistence on minimalism and add binary products to the EL, we could use them to implement our desired semantics like this:
 \begin{lstlisting}
@@
 
-Even if we were to automate this (using, for example, \emph{type macros} as found in Scala \cite{ScalaMacros2013}, adapted to a language with modules), another issue is that we can only reasonably express the introductory and projection operators by specifying their semantics as functions like this.
+Even if we were to automate this (using, for example, \emph{type macros} as  in Scala \cite{ScalaMacros2013}, adapted to a language with ML-style modules), another issue is that we can only reasonably express the introductory and projection operators by specifying their semantics as functions like this.
 To enumerate all possible operators that arise from the opcon \lstinline{ltuple-} using functions would require constructing an encoding of every labeled tuple type that might arise from dropping one or more components from the labeled tuple in question.
@@
 Unlike labeled tuple types, record types are identified up to component reordering, so this mapping does not preserve certain type disequalities between labeled tuple types.
-It is not  possible to allow clients to omit component labels and rely on the positional information available in the type when introducing a value of record type for this reason.\todo{clarify}%, because this information is not maintained by the type system.
+Without positional information available in the type, it is not  possible to allow clients to omit component labels when introducing a value of record type.%, because this information is not maintained by the type system.
 
@@
 \begin{lstlisting}[numbers=none]
-fun Paper_of_tpl (title : title_t, conf : conf_t) => 
-  {title => title, conf => conf}
-fun tpl_of_Paper {title => title, conf => conf} =>
-  (title, conf)
+fun Paper_of_tpl (t : title_t, c : conf_t) => 
+  {title => t, conf => c}
+fun tpl_of_Paper {title => t, conf => c} =>
+  (t, c)
 \end{lstlisting}
@@
 Clients must also bear the syntactic and dynamic costs of explicitly invoking these functions.
-We could define a TSL for every such type to reduce this cost to clients, albeit at an additional cost to providers.
+We could define a TSM for every such type to reduce this cost to clients, albeit at an additional cost to providers.
 %This also means that we cannot omit the labels and rely on positional information when introducing a value of a record type in an analytic position, as we did above with type \lstinline{Paper}.
@@
 \begin{lstlisting}[numbers=none]
-fun Paper_of_tpl (title : title_t, conf : conf_t) => {
-  __0_title => title,
-  __1_conf => conf
-}
-fun tpl_of_Paper {__0_title => title, __1_conf => conf} => 
-  (title, conf)
-fun Paper_of_rcd {title => title, conf => conf} => {
-  __0_title => title,
-  __1_conf => conf
-}
-fun rcd_of_Paper {__0_title => title, __1_conf => conf} = {
-  title => title,
-  conf => conf
+fun Paper_of_tpl (t : title_t, c : conf_t) => {
+  __0_title => t,
+  __1_conf => c
+}
+fun tpl_of_Paper {__0_title => t, __1_conf => c} => 
+  (t, c)
+fun Paper_of_rcd {title => t, conf => c} => {
+  __0_title => t,
+  __1_conf => c
+}
+fun rcd_of_Paper {__0_title => t, __1_conf => c} = {
+  title => t,
+  conf => c
 }
@@
 %This treats membership of a string in a specified regular language exclusively as a static verification condition.
-Such  systems typically rely on refinement specifications supplied in comments:
+Such  systems, e.g. SML CIDRE \cite{Davies:1997:RTC}, typically rely on refinement specifications that are supplied in comments:
 \begin{lstlisting}[numbers=none]
@@
 \end{lstlisting}
-One immediate limitation of this approach is that there is now no way to express the group projection operator described above.
-We cannot define a function \lstinline{#group} that can be used like this:
+One immediate limitation of this approach is that there is now no way to express the group projection operator described above, i.e.
+we cannot define a function \lstinline{#group} that can be applied like this:
 \begin{lstlisting}[numbers=none]
@@
 Another perhaps more useful optimization that is not possible is to represent regular strings that have captured groups in a manner that ensures that group projection has constant dynamic cost (by running the match once when the string is introduced and caching the group values).
-
-Put another way, type refinements are useful when one needs to make ``behavioral'' distinctions within a type, but they are not appropriate when new ``structural'' distinctions are needed.
 Another example where this is relevant is when a language designer considers omitting a primitive type of ``non-negative integers representable in 32 bits'' because this seems ``merely'' to be a refinement of (mathematical) integers.
-But by eliminating the structural distinction, the compiler loses the ability to use a machine representation more suitable to this particular static invariant.
-Of course, structural distinctions are sometimes inconvenient because they require defining and applying coercions between distinct types, but it is certainly not the case that a language can reasonably predict all possible structural distinctions that might be useful \emph{a priori}.
+But by eliminating the ``structural'' distinction between the two types, compilers lose the ability to use a machine representation more suitable to this particular static invariant.
+Of course, such structural distinctions  are sometimes inconvenient because they require defining and applying coercions with possibly non-zero dynamic cost to go from values of one type to another.
+Nevertheless, it is certainly not the case that a language can reasonably predict all possible structural distinctions that might be useful \emph{a priori}.
+
+%Put another way, type refinements are quite useful when one needs only to make ``behavioral'' distinctions within a type, but they are not appropriate when new type disequalities would be useful or new operators need to be introduced.
 %Refinement types cannot be used for this purpose without the possibility of conflict.
@@
 
-Verse introduces a \emph{metamodule system} that gives library providers more direct control over the semantics of the EL.
+Verse introduces a \emph{metamodule system} that gives library providers more direct control over the type structure of the EL.
 Using the metamodule system, library providers can express all of the types and operators discussed above.
@@
 This static function programmatically generates the type's translation given its type parameter.
-Type translations are represented as static values of sort \lstinline{ITy}, which we assume supports standard \emph{quasiquotation} syntax.
+Type translations are represented as static values of sort \lstinline{ITy}, which we assume supports standard \emph{quasiquotation} syntax (e.g.
+as in Scala \cite{shabalin2013quasiquotes} or Lisp \cite{Bawd99a}).
 
@@
 We have chosen here to translate labeled tuple types to nested binary product types internally.
-The type translator generates these by folding over the component mapping (using helper function \lstinline{list_of_comp_spec :: ComponentSpec -> list(Lbl * Ty)}, not shown).
+The type translator generates these by folding over the component mapping (using helper function \lstinline{comp_spec_to_tuples :: ComponentSpec -> list(Lbl * Ty)}, not shown).
 It generates the representation of \lstinline{unit} for the empty case, and nests the binary product types in the recursive case (using the standard \emph{unquote} syntax, \lstinline{%}).
@@
 \begin{lstlisting}[numbers=none]
-fun make_paper(conf : conf_t, title : title_t) : Paper => 
-  Ltuple.intro_unlabeled () (conf; title)
+fun make_paper(c : conf_t, t : title_t) : Paper => 
+  Ltuple.intro_unlabeled () (c; t)
 \end{lstlisting}
@@
 \begin{lstlisting}[numbers=none]
-fun make_paper_tsm(conf : conf_t, title : title_t) : Paper => 
-  ltpl Ltuple Paper_tyidx {confSCSS, ECSStitle}
+fun make_paper_tsm(c : conf_t, t : title_t) : Paper => 
+  ltpl Ltuple Paper_tyidx {cSCSS, ECSSt}
 \end{lstlisting}
@@
 \begin{lstlisting}[numbers=none]
-fun make_paper_tsl(conf : conf_t, title : title_t) : Paper => 
-  {confSCSS, ECSStitle}
+fun make_paper_tsl(c : conf_t, t : title_t) : Paper => 
+  {cSCSS, ECSSt}
 \end{lstlisting}
@@
 Internal expressions are represented as static values of sort \lstinline{IExp}, and again we assume that we can use standard quasiquote syntax.
+The remaining opcons (which we will not consider in detail here) are synthetic opcons, which means they can be used anywhere.
+Rather than taking a type as input, they produce a pair of a type and a translation as output.
 
@@
 To validate it, we assume that the type of each variable is the abstracted type translation of the type the corresponding argument was analyzed against.
-So in this case, the abstracted translation is $(x_0, (x_1, ()))$ and it will be validated under a context where $x_0 : \alpha_0$ and  $x_1 : \alpha_1$ against the abstracted type translation for the type the operation as a whole was being analyzed against, i.e.
-$\alpha_0 \times (\alpha_1 \times \mathsf{unit})$, as discussed above.
-Only after this succeeds are the actual argument translations substituted in to generate the final operation translation, here $({conf}, ({title}, ())$.
+So in this case, the abstracted translation is $(x_0, (x_1, ()))$ and it will be validated against the abstracted type translation for \lstinline{Paper}, i.e.
+$\alpha_0 \times (\alpha_1 \times \mathsf{unit})$, under a context where $x_0 : \alpha_0$ and  $x_1 : \alpha_1$ (this will, of course, succeed).
+Only after this succeeds are the actual argument translations substituted in to generate the final operation translation, here $({c}, ({t}, ())$.
 
@@
 For example, even if regular strings are implemented (by some other metamodule, not shown) as strings satisfying the regular string invariant, \lstinline{Ltuple} knows only that \emph{there exists} some translation for each regular string type, but nothing more.
-If the regular string metamodule later decides to change how it implements regular string types, this would require only local changes to that metamodule.
+Thus, it cannot define ``bad'' opcons like this:
+
+\begin{lstlisting}[numbers=none]
+syn opcon bad {
+	static fun (op_param :: unit, args :: list(Arg)) :: Ty * IExp => 
+	  (rstring /.+/, `""`)
+}
+\end{lstlisting}
+Here, the opcon ignores the arguments and synthesizes a regular string type that the client will assume classifies non-empty strings, yet the corresponding translation is in fact the empty string.
+That is, even if all the opcons defined in the metamodule implementing regular strings maintained this regular string invariant, this opcon would violate it.
+Note however that it is not simply a problem of the translation being well-typed.
+The operator translation (the empty string) is consistent with the type translation (\lstinline{string}), so simply validating well-typedness would not be enough.
+Fortunately, by holding the type translation of regular strings abstract when validating the output of this opcon, the problem is caught -- the empty string is not of type $\alpha$.
+Put another way, the problem with this opcon is that it is not defined in a translation-independent manner with respect to tycons defined in another metamodule.
+This is exactly analagous to the representation independence property underlying modular reasoning with ML-style modules.
+By leveraging type abstraction, we are able to achieve modularity guarantees without requiring mechanized proofs of translation invariants (just as ML does not require mechanized proofs of representation invariants).
+
+%If the regular string metamodule later decides to change how it implements regular string types, this would require only local changes to that metamodule.
 We call this property \emph{translation independence}, by analogy to the \emph{representation independence} property that underlies modular reasoning with ML modules.
-An important point is that we are able to achieve modularity guarantees without requiring mechanized proofs of translation invariants (just as ML does not require mechanized proofs of representation invariants).
 %Translation independence allows library providers to reason locally about \emph{translation invariants} maintained by the operators in a metamodule.
@@
 It would require too much space to give a comprehensive formal account of this mechanism in this proposal, so  we leave this as work that remains to be done (though see below for progress).
-In broad strokes, it is similar to the rule for type translation described above.
-
+In broad strokes, it is similar to the rule for type translation described above in that we validate the operation translation before applying the substitutions for type and expression variables.
+
+Note that the static functions that define these opcons contain code that looks just like the code that would appear in a standard implementation of the first stage of a compiler for a language that built in labeled tuple types.
+The novelty is not in these definitions, but in how they are organized and brought into the language.
 % \begin{mathpar}
@@
 
-The remaining opcons are synthetic opcons, which means they can be used anywhere.
-They operate similarly, but rather than taking a type as input, they produce a pair of a type and a translation as output.
-In every other way, these are analagous to analytic opcons, so we omit the details here.
-Instead, note that the static functions that define these opcons contain code that looks just like the code that would appear in a standard implementation of the first stage of a compiler for a language that built in labeled tuple types.
-The novelty is not in these definitions, but in how they are organized and brought into the language.
-
 \subsubsection{Limitations}
@@
 On the other hand, constructs like ML-style exceptions, Ocaml-style open datatypes or Java-style class-based object systems would not be possible because they require statically tracking which constructors/classes are in scope.
-We plan to discuss adding support for new typing contexts as an avenue for future work in the thesis.
-
-Establishing the metatheory of metamodules is also somewhat tricky, again because it involves reasoning about expressions that are not clearly subexpressions of those in the conclusion of a rule.
+We plan to discuss adding support for new user-defined static contexts as an avenue for future work in the dissertation.
+
+Establishing the metatheory of metamodules is also somewhat tricky, again because it involves reasoning about expressions that are not clearly subexpressions of those in the conclusion of the relevant rules.
 We plan to give an induction principle by which we can prove the necessary metatheorems in the dissertation.
@@
   	(* type translation generator: *)
-  	static fn (param :: ComponentSpec) => fold (list_of_comp_spec param)
+  	static fn (param :: ComponentSpec) :: ITy => fold (comp_spec_to_tuples param)
       `SCSSunitECSS`
@@
   	(* translate intro operator to nested pairs *)
-  	static fn (ty :: Ty, op_param :: unit, args :: list(Arg)) => tycase(ty) {
-        c(ty_param) => (fold (zipExact (list_of_comp_spec ty_param) args)
+  	static fn (ty :: Ty, op_param :: unit, args :: list(Arg)) :: IExp => tycase(ty) {
+        c(ty_param) => (fold (zipExact (comp_spec_to_tuples ty_param) args)
       	  `SCSS()ECSS`
@@
   ^\label{line:introl-start}^syn opcon intro_labeled {
-  	static fn (op_param :: list(Lbl), args :: list(Arg)) => 
-      (* ... same as intro_unlabeled but reorder first ... *)
+  	static fn (op_param :: list(Lbl), args :: list(Arg)) :: Ty * IExp => 
+      (* ... similar to intro_unlabeled but reorder first ... *)
   }^\label{line:introl-end}^
@@
   ^\label{line:prj-start}^syn opcon # {
-  	static fn (op_param :: Lbl, args :: list(Arg)) =>  
+  	static fn (op_param :: Lbl, args :: list(Arg)) :: Ty * IExp =>  
       (* ... generate the appropriate n-fold projection ... *)
@@
   ^\label{line:ltpl-conc-start}^syn opcon + {
-  	static fn (op_param :: unit, args :: list(Arg)) => 
+  	static fn (op_param :: unit, args :: list(Arg)) :: Ty * IExp => 
       (* generate the new type and combine the two nested tuples *)
@@
   ^\label{line:ltpl-drop-start}^syn opcon - {
-  	static fn (op_param : Lbl, args : list(Arg)) => 
-      (* generate the new type and drop the appropriate element *)
+  	static fn (op_param :: Lbl, args :: list(Arg)) :: Ty * IExp => 
+      (* generate the new type and drop the appropriate component *)
   }^\label{line:ltpl-drop-end}^
-} with syntax ltpl for c ^\label{line:ltplx-annotation}^
+} with syntax ltpl^\label{line:ltplx-annotation}^
 (* synonyms used in Sec. 5.1 *)

### 2015-06-18 thesis-proposal/problem.tex
Near final draft of proposal after JA's comments

@@
 Instead, its users are able to modularly  express a wide variety of useful constructs in libraries, using the comparatively small set of primitives that the language builds in.
-Functional programming languages like Standard ML (SML), OCaml and Haskell descend from mathematically elegant typed lambda calculi, diverging mainly in that they consider various pragmatic issues alongside  fundamental metatheoretic issues in the design of their syntax and semantics.
-For example, they build in record types, generalizing the nullary and binary product types more common in minimal calculi, because explicitly labeled components can be cognitively useful to programmers.
-Similarly, they build in derived concrete syntax for common library constructs, like lists.% concision and readability.
+Functional programming languages like Standard ML (SML), OCaml and Haskell descend from mathematically elegant typed lambda calculi, diverging mainly in that they consider various pragmatic issues alongside  essential metatheoretic issues in their design.
+For example, they build in record types, generalizing the nullary and binary product types more common in minimal calculi, because explicitly labeled components can be cognitively useful to human programmers.
+Similarly, they build in derived concrete syntax for some common library constructs, like lists.% concision and readability.
 %Minimal typed lambda calculi play a central role in the study of programming languages, and serve as direct inspiration for  functional programming languages like ML.
@@
 %Tools for  constructing new ``domain-specific'' dialects also continue to proliferate. % The situation is similar across all major language lineages. 
-Indeed, tools that assist with the construction of so-called  ``domain-specific'' language dialects are only becoming increasingly popular. 
+Indeed, tools that assist with the construction of so-called  ``domain-specific'' language dialects are only becoming increasingly popular.\footnote{We consider only so-called ``external DSLs'' here.
+``Internal DSL'' refers to a programming interface that  ``resembles'' that of a distinct dialect, though it is actually defined as a library \cite{fowler2010domain}.}
 {This calls for an investigation}: why is it that programmers and researchers are so often unable to satisfyingly express the constructs that they need in libraries, as modes of use of the ``general-purpose'' primitives already available in major languages today?
@@
 To decrease syntactic cost, library providers construct \emph{derived syntactic dialects}, i.e.
-dialects that are specified by context-independent elaboration to the existing language.
+dialects that are specified by a purely syntactic (i.e.
+context-independent) elaboration to the existing language.
 For example, Ur/Web is a  derived syntactic dialect of Ur (a language that itself descends from ML) that builds in derived syntax (colloquially, ``syntactic sugar'') for SQL queries, HTML elements and other datatypes used for web programming \cite{conf/popl/Chlipala15}.
@@
 derived list syntax in SML \cite{harper1997programming,mthm97-for-dart}.
-Tools like Camlp4 \cite{ocaml-manual} and Sugar* \cite{erdweg2011sugarj,erdweg2013framework} have lowered the engineering costs of constructing derived syntactic dialects in situations like these, contributing to their ongoing proliferation.
+Tools like Camlp4 \cite{ocaml-manual}, Sugar* \cite{erdweg2011sugarj,erdweg2013framework} and the tools around Racket \cite{Flatt:2012:CLR:2063176.2063195} have lowered the engineering costs of constructing derived syntactic dialects in situations like these, contributing to their ongoing proliferation.
 %(we will consider them further in Sec.
@@
 In the most general setting where the dialects in question might be specified by judgements of arbitrary form, this is not a well-defined notion.
-Even if we restrict our interest to dialects specified using formalisms that do operationalize some notion of dialect combination, there is generally no guarantee that the combined dialect will conserve syntactic and semantic properties that can be established about the dialects in isolation.
+Even if we restrict our interest to dialects specified using formalisms that do operationalize this notion of dialect combination, there is generally no guarantee that the combined dialect will conserve syntactic and semantic properties that can be established about the dialects in isolation.
 %In other words, any putative ``combined language'' must formally be considered a  distinct system for which one must derive essentially all metatheorems of interest anew, guided only informally by those derived for the dialects individually.
@@
 %This approach is antithetical to the ideal of a truly \emph{general-purpose language} described at the beginning of this section.
-\newpage
+%\newpage
 \begin{quote}\textit{The recent development of programming languages suggests that the simul\-taneous achievement of simplicity 
@@
 Our broad aim in the work being proposed is to introduce a new functional programming language called Verse\footnote{We distinguish Verse from Wyvern, which is the language referred to in prior publications about some of the work being proposed here, because Wyvern is a group effort evolving independently in some important ways.} that takes more radical steps down the second path just described by giving library providers the ability to directly express new derived concrete syntax as well as new type structure of a variety of designs in a safe and modularly composable manner atop a minimal type-theoretic core.
-The result is that 1) Verse is smaller than comparable languages like ML and Scala, and 2) dialect formation is less often necessary.%This eliminates the needs to build in fewer \emph{ad hoc} constructs and dialects are less frequently necessary.
+As a result,  1) Verse will be smaller than comparable languages like ML and Scala, and 2) dialect formation will be less necessary.%This eliminates the needs to build in fewer \emph{ad hoc} constructs and dialects are less frequently necessary.
 %thereby reducing the need for language dialects and revisions.
@@
 \ref{sec:syntax}, eliminate the need to primitively build in derived concrete syntax specific to library constructs, e.g.
-list syntax as in SML or XML syntax as in Scala and Ur/Web, by giving library providers static control (at a specified type or parameterized family of types) over the parsing and elaboration of delimited segments of concrete syntax.
+list syntax as in SML or XML syntax as in Scala and Ur/Web, by giving library providers static control over the parsing and elaboration of delimited segments of concrete syntax (at a specified type or parameterized family of types).
 %We show how TSMs can be invoked explicitly by clients, or, implicitly based on local type inference.
@@
 \end{enumerate} 
-Both TSMs and metamodules are novel forms of \emph{static code generation} (also often called \emph{static} or \emph{compile-time metaprogramming}), i.e.
-the relevant rules in the static semantics call for the evaluation of \emph{static functions} that generate static representations of expressions and types.
+Both TSMs and metamodules involve \emph{static code generation} (also called \emph{static} or \emph{compile-time metaprogramming}), because the relevant rules in the static semantics call for the evaluation of \emph{static functions} that generate static representations of expressions and types.
 Library providers write these static functions in the Verse \emph{static language} (SL).
@@
 A functional programming language can give library providers the ability to %meta\-pro\-gram\-matic\-ally 
-implement new derived concrete syntax and type structure atop a minimal type-theoretic internal language while maintaining a hygienic type discipline and modular reasoning principles.
+express new derived concrete syntax and type structure atop a minimal type-theoretic internal language while maintaining a hygienic type discipline and modular reasoning principles.
 %These  primitives are  expressive enough to subsume the need for a variety of primitives that are, or would need to be, built in to comparable contemporary languages.
@@
 We  expect that in practice, Verse will come with a standard library defining a carefully curated collection of standard constructs, as well as guidelines for advanced users regarding when it would be sensible to use the mechanisms we introduce (following the example of languages that support operator overloading or type classes \cite{Hall:1996:TCH:227699.227700}, which also have the potential for such ``abuse'').
-For most programmers, using Verse should not be substantially different from using a comparable language or one of its dialects.%The vast majority of programmers should not use the primitives that we introduce directly.
-
-Finally, Verse intentionally does not support dependent types ala Coq, Agda or Idris because they do not maintain a phase separation between ``compile-time'' and ``run-time.'' This phase separation is useful for programming tasks (where one would like to be able to discover errors before running a program, particularly ones that may have an effect) but less so for theorem provers (where it is mainly the fact that a pure expression is well-typed that is of interest).
-Verse is designed for use for programming tasks where SML, Ocaml, Haskell or Scala would be used today, not for sophisticated theorem proving.
+For most programmers, using Verse should not be substantially different from using a language like ML or one of its dialects.%The vast majority of programmers should not use the primitives that we introduce directly.
+
+Finally, Verse intentionally is not a dependently-typed language like Coq, Agda or Idris, because these languages do not maintain a phase separation between ``compile-time'' and ``run-time.'' This phase separation is useful for programming tasks (where one would like to be able to discover errors before running a program, particularly programs that may have an effect) but less so for theorem proving tasks (where it is mainly the fact that a pure expression is well-typed that is of interest, by the propositions-as-types principle).
+Verse is designed to be used for programming tasks where SML, Ocaml, Haskell or Scala would be used today, not for advanced theorem proving tasks.
 That said, we conjecture that the primitives we describe could be added to languages like Gallina (the ``external language'' of the Coq proof assistant  \cite{Coq:manual}) or to the ``program extraction'' mechanisms of proof assistants like Coq with  modifications, but do not plan to pursue this line of research in this dissertation.
@@
 The key novelty is that the core language is split into a user-facing \emph{typed external language} (EL) specified by type-directed translation to a minimal \emph{typed internal language} (IL).
-The core language can notionally be compared to the first stage of a type-directed compiler (e.g.
-the TIL compiler for Standard ML \cite{tarditi+:til-OLD}) shifted ``one level up'' into the language itself.
+One might think of this design as shifting the first stage of a type-directed compiler (e.g.
+the TIL compiler for Standard ML \cite{tarditi+:til-OLD})  ``one level up'' into the language itself.
 A third sublanguage, the \emph{typed static language} (SL), is involved in giving library providers programmatic control over aspects of this type-directed translation from the EL to the IL.
 % The SL is notionally analagous to the language that the first stage itself is written in.
-
 We summarize the main judgements relevant to the IL, EL and SL below. Readers who prefer to see examples first can safely skim these sections for now.
@@
 
-Note that features like state, exceptions, minimal concurrency primitives, scalars, arrays and other primitives characteristic of a first-stage compiler intermediate language would also be included in the IL in practice, and for several of these, this would affect the shape of the internal semantics.
+Note that features like state, exceptions, basic concurrency primitives, scalars, arrays, I/O primitives and others characteristic of a first-stage compiler intermediate language would also be included in the IL in practice, and for several of these, this would affect the shape of the internal semantics.
 However, our design is largely insensitive to such details -- the only strict requirements are that the IL be type safe and support parametric type abstraction.
-We wish to give a minimal account of our novel contributions, so we will stick to this simpler, well-understood IL for the purposes of this work.
+To give a simple account of our novel contributions, so we will stick to this small, familiar IL for the purposes of this work.
 %Features like record types and labeled sum types, on the other hand, are not built in.
@@
 The external language supports a richer syntax and type structure atop the IL.
-Throughout this work, the words ``expression'' and ``type'' used without qualification refer to external expressions and types, respectively.
-The main judgements in the specification of the EL take the following form  (again omitting various contexts for now):
+Most programmers will interact exclusively with the EL, so throughout this work, the words ``expression'' and ``type'' used without qualification refer to external expressions and types, respectively.
+The main judgements in the specification of the EL take the following form  (again omitting various contexts, which we will introduce progressively):
 \\[1ex]
@@
 \subsubsection{Static Language}
-Finally, the static language plays an essential role in the semantics of the novel primitives that we will introduce into the EL.
-The SL is another typed lambda calculus consisting of \emph{static expressions}, $\sigma$, classified by \emph{sorts}\footnote{We use this word to avoid confusion with the phrase ``static type''.}, $\uppi$, according to the following judgements (the ``static statics''):
+Finally, the static language plays an essential role in the semantics of the novel primitives that are the topic of this proposal.
+The SL is another typed lambda calculus consisting of \emph{static expressions}, $\sigma$, classified by \emph{sorts}\footnote{We use the word ``sort'' to avoid confusion with the phrase ``static type''.
+Our use of the  word ``sort'' should be considered distinct from the various other uses of this word found in the literature on programming languages.}, $\uppi$, according to the following judgements (the ``static statics''):
 \\[1ex]
@@
 
-As suggested by the form of the external type translation judgement above, external types are static values of a primitive sort written $\mathsf{Ty}$.
-We will return to how this works (and consider the metatheoretic implications of this choice) in Sec.
+As suggested by the form of the external type translation judgement above, external types are static expressions.
+More specifically, external types are static values of a primitive sort written $\mathsf{Ty}$.
+We will return to how this works in Sec.
 \ref{sec:metamodules}.
@@
 \subsection{Module Language}
-We will first specify our contributions assuming a language without an ML-style module system, then make modifications to this specification to make it compatible with the addition of a module system.
-
-The Verse module language is taken directly from Standard ML, with which we assume a working familiarity for the purposes of this proposal \cite{harper1997programming,MacQueen:1984:MSM:800055.802036} (a related module language, e.g.
+
+The Verse module system is taken directly from Standard ML, with which we assume a working familiarity for the purposes of this proposal \cite{harper1997programming,MacQueen:1984:MSM:800055.802036} (a related module system, e.g.
 Ocaml's, would work just as well for our purposes).
-We will give examples of its use in Sections \ref{sec:syntax} and \ref{sec:metamodules}, but because it has been thoroughly studied in the literature, we will defer to prior work both here and in the dissertation for most formal details.
-In particular, W In doing so, we will assume that modules are being tracked by a module context, without detailing how this context is populated.
+We will give examples of its use in Sec.
+\ref{sec:syntax} and Sec.
+\ref{sec:metamodules}.
+
+Formally, our approach will be to first specify our contributions assuming a language without an ML-style module system, then make additions to this specification to make it compatible with the presence of a module system.
+Because it has been thoroughly studied in the literature, we will defer to prior work both here and in the dissertation for many formal details.
+In particular, we will assume that modules are being tracked by a suitable \emph{module context}, without detailing how this context is populated.
 %This context will be analagous to the \emph{named type context} specified in the papers just cited.% Instead, we will consider it abstractly where it interacts with the core language.%\todo{reminder that we are not going to re-specify the module system}
@@
 \section{Modularly Programmable Textual Syntax}\label{sec:syntax}
-To begin, let us assume a standard, ML-like semantics for the EL and SL and consider the EL's concrete syntax.
-Verse, like many major contemporary languages, specifies a textual concrete syntax for the EL.\footnote{Although Wyvern specified a layout-sensitive concrete syntax, to avoid unnecessary distractions, we will specify a more conventional layout-insensitive concrete syntax for Verse.} %We have chosen to specify a layout-sensitive textual concrete syntax (i.e.
+To begin, let us assume a standard, ML-like semantics for both the EL and SL and consider the EL's concrete syntax.
+Verse, like most major contemporary languages, specifies a textual concrete syntax.\footnote{Although Wyvern specified a layout-sensitive concrete syntax, to avoid unnecessary distractions, we will specify a more conventional layout-insensitive concrete syntax for Verse.} %We have chosen to specify a layout-sensitive textual concrete syntax (i.e.
 newlines and indentation are not ignored).
@@
 %For constructs that have an obvious analog in ML, we will omit a detailed explanation.
-Because the purpose of concrete syntax is to serve as the programmer-facing user interface of the language, it is common practice to build in  derived syntactic forms (colloquially, \emph{syntactic sugar}) that capture common idioms more concisely or naturally.
+Because the purpose of this syntax is to serve as the programmer-facing user interface of the language, it is common practice to build in  derived syntactic forms (colloquially, \emph{syntactic sugar}) that capture common idioms more concisely or naturally.
 % (i.e.
@@
 
-Verse takes a less \emph{ad hoc} approach -- rather than privileging  particular library constructs with primitive syntactic support, Verse exposes primitives that allow library providers to introduce new derived syntax of a variety of designs on their own, in a safe and modular manner.
+Verse takes a less \emph{ad hoc} approach -- rather than privileging  particular library constructs with primitive syntactic support, Verse exposes primitives that allow library providers to introduce new elaboration logic on their own, in a safe and modular manner.
 %Lists need no special consideration from the language specification.
@@
 %For forms with a clear analogy to a form in Standard ML, we will assume the  semantics are analagous without providing details.
-We begin in Sec.
+
+We will begin in Sec.
 \ref{sec:examples} by detailing a representative example for which such a mechanism would be useful: regular expression patterns expressed using abstract data types.
@@
 In Sec.
-\ref{sec:syntax-contributions}, we outline our proposed alternatives -- \emph{typed syntax macros} and the related \emph{type-specific languages} -- and discuss how they resolve these issues.
-We also discuss how TSMs are specified in terms of the judgement forms introduced in the previous section.
+\ref{sec:syntax-contributions}, we outline our proposed alternatives -- \emph{typed syntax macros} (TSMs) and the related \emph{type-specific languages} (TSLs) -- and discuss how they resolve these issues.
+We also show how TSMs are specified in terms of the judgement forms that were introduced in the previous section.
 We conclude in Sec.
@@
 The identity of this type is held abstract outside the module during typechecking (i.e.
-it acts as a newly generated type) so that the burden of proving that there is no way to use the case analysis function to distinguish patterns with the same normal form is local to the module, and implementation details do not escape (and can thus change freely).
+it acts as a newly generated type) so that the burden of proving that there is no way to use the case analysis function to distinguish patterns with the same normal form is local to the module, and implementation details do not escape (and can thus evolve freely).
 %The details are standard and not particularly relevant for our purposes, so we omit them here.
@@
 \end{lstlisting}
-Though ill-formed patterns can sometimes be discovered dynamically via testing, empirical data gathered from large open source projects suggests that there remain many ill-formed regular expression patterns that are not detected by a project's test suite ``in the wild'' \cite{spishak2012type}.
-
-Statically verifying that pattern formation errors will not dynamically arise is quite difficult to automate in general, because it requires reasoning about the full dynamics of the language.
-In this example, the verification procedure would need to be able to establish that the variable \lstinline{pattern} is equal to the parse function \lstinline{PU.parse}.
-If the string argument had not been written literally but rather computed, e.g.
+Though malformed patterns can sometimes be discovered dynamically via testing, empirical data gathered from large open source projects suggests that there remain many malformed regular expression patterns that are not detected by a project's test suite ``in the wild'' \cite{spishak2012type}.
+
+Statically verifying that pattern formation errors will not dynamically arise requires reasoning about the full dynamics of the language.
+This is an undecidable verification problem in general and is difficult to even partially automate.
+In this example, the verification procedure would first need to be able to establish that the variable \lstinline{pattern} is equal to the parse function \lstinline{PU.parse}.
+If the string argument had not been written canonically but rather computed, e.g.
 as \lstinline{"SSTR(GESTR" ^ "SSTRCESTR"} where \lstinline{^} is the string concatenation function applied in infix style, it would also need to be able to establish that this expression is equivalent to the string \lstinline{"SSTR(GCESTR"}.
-For patterns that are dynamically constructed based on input to a function (see below), the problem is not solved by ``simply''  evaluating the expression statically (or, more generally, in some earlier ``stage'' of evaluation) \cite{Jones:Gomard:Sestoft:93:PartialEvaluation}.
-
-Of course, asking the client to provide a proof of well-formedness would completely defeat our purpose -- providing a concise concrete syntax.
+For patterns that are dynamically constructed based on input to a function (see example below), the problem is not solved by ``simply''  evaluating the expression statically (or, more generally, in some earlier ``stage'' of evaluation) \cite{Jones:Gomard:Sestoft:93:PartialEvaluation}.
+
+Of course, asking the client to provide a proof of well-formedness would completely defeat the purpose -- lowering syntactic cost.
 
@@
 To avoid incurring the parsing cost for each element of \lstinline{exmpl_list}, the programmer or compiler must move the parsing step out of the closure or, barring that, ensure that an appropriately tuned memoization (i.e.
-caching) strategy is being used.\footnote{Anecdotally, in scientific code using standard contemporary compilers, this optimization is not automatic.
+caching) strategy is being used.\footnote{Anecdotally, in scientific code using major contemporary compilers, the former optimization is not automatic.
 Compilation caches are more often provided by the regular expression library, but like any cache, it is difficult to reason compositionally about performance in their presence.} %If the programmer does it, it can sometimes make the program more difficult to read.
@@
 
-\item The next problem is that dynamic string parsing mainly decreases the syntactic cost of statically known patterns.
-Patterns constructed dynamically cannot easily benefit from this technique.
+\item The next problem is that dynamic string parsing mainly decreases the syntactic cost of ``literal'' patterns.
+Patterns constructed compositionally cannot easily benefit from this technique.
 For example, consider this function from strings to patterns:
@@
 It is only when the input \lstinline{name} contains special characters that have meaning in the concrete syntax of patterns that a problem arises.
-In applications that query sensitive data, mistakes like this lead to \emph{injection attacks}, which are among the most common and catastrophic security threats today \cite{owasp2013}.
-Proving that mistakes like these have not been made involves reasoning about complex run-time data flows, so it is once again notoriously difficult to automate.
+In applications that query sensitive data, mistakes like this lead to \emph{injection attacks}, which are among the most common and catastrophic security threats on the web today \cite{owasp2013}.
+Proving that mistakes like this have not been made involves reasoning about complex run-time data flows, so it is once again notoriously difficult to automate.
  %Ultimately, of course, mistakes like this are the fault of a programmer using a flawed heuristic, and they could be avoided with discipline.
@@
 A formal account of such a system has been developed by Griffin \cite{5134}.
-Unfortunately, these systems are not flexible enough to allow us to express pattern syntax in the conventional manner.
+Unfortunately, these systems are not expressive enough to allow us to express pattern syntax in the conventional manner.
 For example, sequences of characters  can only be parsed as identifiers using these systems, rather than as characters in a regular expression pattern.
 
-Syntax extension systems based on context-free grammars like  Sugar* \cite{erdweg2013framework}, Camlp4 \cite{ocaml-manual} and several others are more expressive, and would allow us to directly introduce pattern syntax into our core language's grammar, perhaps following Unix conventions like this:
+Syntax extension systems based on context-free grammars like  Sugar* \cite{erdweg2013framework}, Camlp4 \cite{ocaml-manual} and several others, as well as mechanisms that allow direct programmatic control over parsing like Racket's reader extensions \cite{Flatt:2012:CLR:2063176.2063195} are more expressive, and would allow us to directly introduce pattern syntax into our core language's grammar, perhaps following Unix conventions like this:
 \begin{lstlisting}[numbers=none]
@@
 
-For patterns constructed dynamically, we can define \emph{splicing syntax} that allows us to write string and pattern expressions inline (distinguished by prefixes \lstinline{@} and \lstinline{%}, respectively).
-For example, we can write \lstinline{example} equivalently, but at much lower syntactic cost, like this:
+For patterns constructed compositionally, we can define \emph{splicing syntax} that allows us to write string and pattern expressions inline (distinguished by prefixes \lstinline{@} and \lstinline{%}, respectively).
+For example, we can write \lstinline{example} equivalently, but at lower syntactic cost, like this:
 \begin{lstlisting}[numbers=none,escapechar=|]
@@
 
-All of the problems of dynamic string parsing described above are solved by this approach.
-Unfortunately, it introduces new problems.
+The problems of dynamic string parsing described above are solved by this approach.
+Unfortunately, it introduces serious new problems.
 First, the systems mentioned thus far cannot guarantee that \emph{syntactic conflicts} between such extensions will not arise.
 As stated directly in the  Coq manual: ``mixing different symbolic notations in [the] same text may cause serious parsing ambiguity''.
-If another library provider used similar syntax for a different implementation or variant of regular expressions, or for some other unrelated construct, then a client could not simultaneously use both libraries in the same piece of code.
+If another library provider used similar syntax for a different implementation or variant of regular expressions, or for some other unrelated construct, then a client could not simultaneously use both libraries at the same time.
 So properly considered, every combination of extensions introduced using these mechanisms creates a \emph{de facto} derived syntactic dialect of our language.
@@
 However, note that the most natural starting tokens like \lstinline{pattern} cannot be guaranteed to be universally unique, so we would be forced to use a more verbose token like \lstinline{edu_cmu_verse_rx_pattern}.
-There is no way for clients of our extension to define scoped abbreviations for starting tokens because this mechanism operates purely on syntactic forms and has no notion of semantic scope.
+There is no simple way for clients of our extension to define scoped abbreviations for starting tokens because this mechanism operates purely on syntactic forms.
 
@@
 
-Putting aside this question of hygiene as well, we can address the problem of choosing a module to use in the elaboration by requiring that the client explicitly identify it as an ``argument'' to the form:
+We can address the problem of choosing a module to use in the elaboration by requiring that the client explicitly identify it as an ``argument'' to the form:
 \begin{lstlisting}[numbers=none]
@@
 It statically parses the provided string literal (which must still be written using an Ocaml-style literal here) to generate an elaboration.
-The macro specifies that the elaboration will be of type \lstinline{P.t}.
-By parsing the string statically, we avoid the problems of dynamic string parsing for statically known patterns.
-
-For patterns that are constructed dynamically, we need to get more creative.
+The macro specifies that the elaboration it statically generates will be of type \lstinline{P.t}.
+By parsing the string statically, we avoid the problems of dynamic string parsing for statically-known patterns.
+
+For patterns that are constructed compositionally, we need to get more creative.
 For example, we might repurpose the infix operators that are normally used for other purposes to support string and pattern splicing as follows:
@@
 
-Having to creatively repurpose existing syntax like this limits the improvements a library provider can make to syntactic cost (particularly when there are pre-existing conventions that are not consistent with the conventions adopted by the language).
-This also can create confusion for readers expecting parenthesized expressions to behave in a consistent manner.
-However,  this approach does avoid many of the problems of na\"ive syntax extensions: there cannot be syntactic conflicts (because the syntax is not extended at all), we can define macro abbreviations because macros are integrated into the language, there is a hygiene discipline that guarantees that the elaboration will not capture variables inadvertently, and using a typed macro system, there is a type discipline (i.e.
+Having to creatively repurpose existing syntax like this limits the impact a library provider can have on syntactic cost (particularly when there are pre-existing conventions that are not consistent with the conventions adopted by the language).
+It also can create confusion for readers expecting parenthesized expressions to behave in a consistent manner.
+However,  this approach is still common because it avoids many of the problems of na\"ive syntax extensions: there cannot be syntactic conflicts (because the syntax is not extended at all), we can define macro abbreviations because macros are integrated into the language, there is a hygiene discipline that guarantees that the elaboration will not capture variables inadvertently, and using a typed macro system, there is a clear type discipline (i.e.
 programmers need not examine the elaboration to know what type a macro application must have).
@@
 Verse provides a new external primitive -- the \textbf{typed syntax macro} (TSM) -- that combines the syntactic flexibility of syntax extensions with the reasoning guarantees of typed macros.
-TSMs can be parameterized by modules, so they can be used to define syntax valid at all types defined in modules satisfying a specified signature.
-As we will discuss in in Sec.
+TSMs can be parameterized by modules, so they can be used to define syntax valid at any type defined by a module satisfying a specified signature.
+As we will discuss in Sec.
 \ref{sec:tsms} below, this addresses all of the problems brought up above, at moderate syntactic cost.
@@
 
-Syntax defined by library providers using TSMs and TSLs comes at syntactic cost comparable to that of derived syntax built in primitively by the language designer or  using a na\"ive syntax extension tool like Camlp4, without resulting in dialect formation.
+Syntax defined by library providers using these primitives comes at syntactic cost comparable to that of derived syntax built in primitively by the language designer or  using a na\"ive syntax extension tool like Camlp4, without causing dialect formation.
 As such, these \emph{ad hoc} approaches can, in large part, be discarded.
@@
 \end{lstlisting}
-Here, a \emph{parameterized TSM}, \lstinline{pattern}, is being applied first to a module parameter, \lstinline{P}, then to a \emph{delimited form}, \lstinline{/SURLA|T|G|CEURL/}.
+Here, we apply a \emph{parameterized TSM}, \lstinline{pattern}, first to a module parameter, \lstinline{P}, then to a \emph{delimited form}, \lstinline{/SURLA|T|G|CEURL/}.
 Note that a number of alternative delimiters are also provided by Verse's concrete syntax and could equivalently be used.
@@
 In practice, this too should  be a minor concern.
-A compiler would be free to limit the running time of the parse function to ensure that compilation does not ``hang'' due to an incorrect parser.
-Related concerns would arise if the static language supported state or access to other external effects.
-Existing techniques for controlling effects would handle this problem, but this is not something we plan to detail here because our SL does not specify these effects.
+For example, we could specify a limit to the running time of the parse function.
+The compiler would enforce this limit to ensure that compilation does not ``hang'' due to an incorrect parser.
+Related concerns would arise if the static language supported external (e.g.
+I/O) effects.
+Existing techniques for controlling these effects would handle this problem, but this is not something we plan to detail here because our SL does not specify these effects.
 
@@
 The mechanics of defining a TSM, statically invoking the parse function, elaboration encoding/decoding and hygienic elaboration validation have all been detailed there.
-In the dissertation, I plan to first present the formal system closely following this presentation, updating it only to ensure that it is consistent with the work in the next section to present a unified design.
+In the dissertation, I plan to first present the formal system closely following this presentation, updating it only to ensure that it is consistent with the description above and with the work in the next section.
 The rule (syn-aptsm) above represents the first and most substantial step of this plan.
@@
 
-Wyvern does not have an ML-style module system (type declarations are the only form of type generation) and at the time, it did not have parameterized types, so to actually support the examples we have given in this section, we also need to update the formalization to handle parameterized TSMs.
+Wyvern does not specify an ML-style module system (type declarations are the only form of type generation) and at the time, it did not support parameterized types, so to actually support the examples we have given in this section, we then need to update the formalization to handle parameterized TSMs.
 I plan to flesh this out in the course of writing the dissertation, and I anticipate it requiring 5.5 weeks of my time after the proposal is complete.
@@
 \ref{sec:metamodules-related}.
-Next, we outline how \emph{metamodules} (together with TSMs and TSLs) serve to address these problems in Sec.
+Next, we outline how {metamodules} (together with TSMs and TSLs) serve to address these problems in Sec.
 \ref{sec:metamodules-solution} and conclude with a timeline in Sec.
@@
 where the expected type is known).
-For example, we can apply \lstinline{Ltuple.intro_unlabeled} like this, because the return type of the function determines the expected type:
+For example, we can apply \lstinline{Ltuple.intro_unlabeled} as follows, because the return type of the function determines the expected type:
 \begin{lstlisting}[numbers=none]
@@
 \end{lstlisting}
-First, we provide the opcon with a static parameter value of the appropriate sort, here \lstinline{unit}, to create an operator.
+First, we provided the opcon with a static parameter value of the appropriate sort, here \lstinline{unit}, to create an operator.
 Then, we supply the operator with an \emph{argument list}, \lstinline{(conf; title)}.
@@
 syn opcon bad {
-	static fun (op_param :: unit, args :: list(Arg)) :: Ty * IExp => 
+	static fn (op_param :: unit, args :: list(Arg)) :: Ty * IExp => 
 	  (rstring /.+/, `""`)
@@
 \end{lstlisting}
-Here, the opcon ignores the arguments and synthesizes a regular string type that the client will assume classifies non-empty strings, yet the corresponding translation is in fact the empty string.
-That is, even if all the opcons defined in the metamodule implementing regular strings maintained this regular string invariant, this opcon would violate it.
-Note however that it is not simply a problem of the translation being well-typed.
-The operator translation (the empty string) is consistent with the type translation (\lstinline{string}), so simply validating well-typedness would not be enough.
+Here, the opcon ignores the arguments and synthesizes a regular string type that the client will assume classifies non-empty strings.
+However the corresponding translation is in fact the empty string.
+So even if all of the opcons defined in the metamodule implementing regular strings maintained this regular string invariant, this opcon would violate it.
+Note  that simply checking the translation is  well-typed would not solve this problem.
+The operator translation (the empty string) is consistent with the type translation (\lstinline{string}).
 Fortunately, by holding the type translation of regular strings abstract when validating the output of this opcon, the problem is caught -- the empty string is not of type $\alpha$.
@@
 
-To formalize the intuition just described, we would need to introduce a number of technical devices.
-It would require too much space to give a comprehensive formal account of this mechanism in this proposal, so  we leave this as work that remains to be done (though see below for progress).
+To formalize the intuition just described, we will need to introduce a non-trivial number of technical devices.
+It would thus require too much space to give a comprehensive formal account of this mechanism in this proposal, so  we leave this as work that remains to be done (though see below for progress).
 In broad strokes, it is similar to the rule for type translation described above in that we validate the operation translation before applying the substitutions for type and expression variables.

### 2015-06-18 thesis-proposal/problem.tex
Fixed some missing things

@@
 % The SL is notionally analagous to the language that the first stage itself is written in.
+
 We summarize the main judgements relevant to the IL, EL and SL below. Readers who prefer to see examples first can safely skim these sections for now.
@@
 Regular string types are constructed by applying the tycon \lstinline{rstring} to a statically valued regular expression pattern, again written here using conventional concrete syntax like that discussed in the previous section.
-Regular string types classify expressions that behave as strings in the corresponding regular language, here the language of non-empty strings.
 The second component of type \lstinline{Paper} is labeled \lstinline{conf} and specifies a regular string type with two {parenthesized groups}, corresponding to a conference abbreviation and year.
@@
 
-We can introduce a value of type \lstinline{Paper} in an analytic position in one of two ways. We can omit the labels and provide component values positionally:
+Clients can introduce regular strings in analytic position using standard string literal syntax, as long as the body of the literal is in the corresponding regular language. 
+Clients can introduce a value of type \lstinline{Paper} in an analytic position in one of two ways. We can omit the labels and provide component values positionally:
 \begin{lstlisting}[numbers=none]
@@
 %This treats membership of a string in a specified regular language exclusively as a static verification condition.
-Such  systems, e.g. SML CIDRE \cite{Davies:1997:RTC}, typically rely on refinement specifications that are supplied in comments:
+Such  systems allow programmers to specify \emph{refinements} of a type that capture more specific invariants.
+For example, we might consider capturing the regular string invariant using refinements of the \lstinline{string} type.
+Clients supply refinement specifications using structured comments or annotations (which can then be processed by external refinement checkers, like SML CIDRE \cite{Davies:1997:RTC}):
 \begin{lstlisting}[numbers=none]

### 2015-06-18 thesis-proposal/problem.tex
Fixed some missing things

@@
 Indeed, tools that assist with the construction of so-called  ``domain-specific'' language dialects are only becoming increasingly popular.\footnote{We consider only so-called ``external DSLs'' here.
-``Internal DSL'' refers to a programming interface that  ``resembles'' that of a distinct dialect, though it is actually defined as a library \cite{fowler2010domain}.}
+The terms ``internal'' and ``embedded DSL'' refer informally to programming interfaces that  ``resemble'' distinct dialects, but that can be expressed within libraries, using existing primitives (often using some creative trick) \cite{fowler2010domain}.}
 {This calls for an investigation}: why is it that programmers and researchers are so often unable to satisfyingly express the constructs that they need in libraries, as modes of use of the ``general-purpose'' primitives already available in major languages today?
@@
 \end{enumerate} 
-Both TSMs and metamodules involve \emph{static code generation} (also called \emph{static} or \emph{compile-time metaprogramming}), because the relevant rules in the static semantics call for the evaluation of \emph{static functions} that generate static representations of expressions and types.
+Both TSMs and metamodules involve \emph{static code generation} (also called \emph{static} or \emph{compile-time metaprogramming}), in that the relevant rules in the static semantics call for the evaluation of \emph{static functions} that generate static representations of expressions and types.
 Library providers write these static functions in the Verse \emph{static language} (SL).
@@
 A third sublanguage, the \emph{typed static language} (SL), is involved in giving library providers programmatic control over aspects of this type-directed translation from the EL to the IL.
-% The SL is notionally analagous to the language that the first stage itself is written in.
-
-We summarize the main judgements relevant to the IL, EL and SL below. Readers who prefer to see examples first can safely skim these sections for now.
+This design has conceptual roots in earlier work on \emph{active libraries}, which similarly envisioned using compile-time computation to give library providers more control over various aspects of the language and compilation process \cite{active-libraries-thesis}.% The SL is notionally analagous to the language that the first stage itself is written in.
+
+Technically, we take a standard type theoretic approach (we will assume that readers are familiar with the early chapters of a textbook like \emph{TAPL} \cite{tapl} or \emph{PFPL} \cite{pfpl}).
+We summarize the main judgements relevant to the IL, EL and SL below.
+Readers who prefer to see examples first can skim these sections for now, returning to them as needed.
 
@@
 Another example where this is relevant is when a language designer considers omitting a primitive type of ``non-negative integers representable in 32 bits'' because this seems ``merely'' to be a refinement of (mathematical) integers.
-But by eliminating the ``structural'' distinction between the two types, compilers lose the ability to use a machine representation more suitable to this particular static invariant.
+But by eliminating the structural distinction between the two types, compilers lose the ability to use a machine representation more suitable to this particular static invariant.
 Of course, such structural distinctions  are sometimes inconvenient because they require defining and applying coercions with possibly non-zero dynamic cost to go from values of one type to another.

### 2015-06-18 thesis-proposal/problem.tex
Fixed some missing things

@@
 ML5 builds in high-level primitives for distributed programming based on a modal lambda calculus \cite{Murphy:2007:TDP:1793574.1793585}.
-Manticore \cite{conf/popl/FluetRRSX07} and AliceML  \cite{AliceLookingGlass} build in parallel programming primitives with a more elaborate type structure than is found in their respective core calculi.
+Manticore \cite{conf/popl/FluetRRSX07} and AliceML  \cite{AliceLookingGlass} build in parallel programming primitives with a more elaborate type structure than is found in core calculi for parallel programming.
 MLj builds in the type structure of the Java object system (motivated by a desire to interface safely and naturally with Java libraries) \cite{Benton:1999:IWW:317636.317791}.

### 2015-06-22 thesis-proposal/problem.tex
Final proposal

@@
 Instead, its users are able to modularly  express a wide variety of useful constructs in libraries, using the comparatively small set of primitives that the language builds in.
-Functional programming languages like Standard ML (SML), OCaml and Haskell descend from mathematically elegant typed lambda calculi, diverging mainly in that they consider various pragmatic issues alongside  essential metatheoretic issues in their design.
-For example, they build in record types, generalizing the nullary and binary product types more common in minimal calculi, because explicitly labeled components can be cognitively useful to human programmers.
-Similarly, they build in derived concrete syntax for some common library constructs, like lists.% concision and readability.
+Functional programming languages like Standard ML (SML), OCaml and Haskell descend from mathematically elegant typed lambda calculi, diverging mainly in that they consider various pragmatic issues in addition to  essential metatheoretic issues in their design.
+For example, they build in record types, generalizing the nullary and binary product types that suffice in simpler calculi, because explicitly labeled components can be cognitively useful to human programmers.
+Similarly, they build in derived syntax (or colloquially, ``syntactic sugar'') for selected library constructs, like lists.% concision and readability.
 %Minimal typed lambda calculi play a central role in the study of programming languages, and serve as direct inspiration for  functional programming languages like ML.
@@
 %syntactic and semantic primitives that capture common idioms naturally and precisely.
-
-The hope is that a limited number of ``embellishments'' like these will suffice to turn a minimal lambda calculus into a broadly practical programming language.
+e
+The hope is that a limited number of primitive ``embellishments'' like these will suffice to turn a minimal lambda calculus into a ``general-purpose'' programming language.
 %The community around this language should, ideally, then be able to direct its efforts almost exclusively toward the development of useful libraries, rather than further modifications to the language itself.
@@
 %Tools for  constructing new ``domain-specific'' dialects also continue to proliferate. % The situation is similar across all major language lineages. 
-Indeed, tools that assist with the construction of so-called  ``domain-specific'' language dialects are only becoming increasingly popular.\footnote{We consider only so-called ``external DSLs'' here.
-The terms ``internal'' and ``embedded DSL'' refer informally to programming interfaces that  ``resemble'' distinct dialects, but that can be expressed within libraries, using existing primitives (often using some creative trick) \cite{fowler2010domain}.}
-{This calls for an investigation}: why is it that programmers and researchers are so often unable to satisfyingly express the constructs that they need in libraries, as modes of use of the ``general-purpose'' primitives already available in major languages today?
+Indeed, tools that assist with the construction of so-called  ``domain-specific'' language (DSL) dialects\footnote{Such dialects are also sometimes called ``external DSLs''.
+The terms ``internal'' and ``embedded DSL'' refer informally to programming interfaces that  ``resemble'' distinct dialects, but that are expressed as libraries, using existing primitives \cite{fowler2010domain}.} are becoming increasingly prominent.
+{This calls for an investigation}: why is it that programmers and researchers are still so often unable to satisfyingly express the constructs that they need in libraries, as modes of use of the ``general-purpose'' primitives already available in major languages today?
 %Let us investigate why this is.
@@
 %Library implementations are not yet satisfying.
-
-Perhaps the most common reason for this proliferation of dialects may simply be that the \emph{syntactic cost} of expressing a construct of interest using contemporary general-purpose primitives is not always ideal.
-To decrease syntactic cost, library providers construct \emph{derived syntactic dialects}, i.e.
-dialects that are specified by a purely syntactic (i.e.
+%\subsection{Why are there so many dialects?}
+Perhaps the most common reason for this ongoing proliferation of dialects may simply be that the \emph{syntactic cost} of expressing a construct of interest using contemporary general-purpose primitives is not always ideal.
+To decrease syntactic cost, library providers often construct \emph{syntactic dialects}, i.e.
+dialects that can be specified by a purely syntactic (i.e.
 context-independent) elaboration to the existing language.
-For example, Ur/Web is a  derived syntactic dialect of Ur (a language that itself descends from ML) that builds in derived syntax (colloquially, ``syntactic sugar'') for SQL queries, HTML elements and other datatypes used for web programming \cite{conf/popl/Chlipala15}.
+For example, Ur/Web is a syntactic dialect of Ur (a language that itself descends from ML) that builds in derived syntax for SQL queries, HTML elements and other datatypes used in the domain of web programming \cite{conf/popl/Chlipala15}.
 %Syntactic cost is often assessed qualitatively \cite{green1996usability}, though quantitative metrics can be defined.
@@
 derived list syntax in SML \cite{harper1997programming,mthm97-for-dart}.
-Tools like Camlp4 \cite{ocaml-manual}, Sugar* \cite{erdweg2011sugarj,erdweg2013framework} and the tools around Racket \cite{Flatt:2012:CLR:2063176.2063195} have lowered the engineering costs of constructing derived syntactic dialects in situations like these, contributing to their ongoing proliferation.
+Tools like Camlp4 \cite{OCaml-manual}, Sugar* \cite{erdweg2011sugarj,erdweg2013framework} and Racket \cite{Flatt:2012:CLR:2063176.2063195} have lowered the engineering costs of constructing syntactic dialects in situations like these, contributing to their ongoing proliferation.
 %(we will consider them further in Sec.
@@
 %The syntactic cost of  such encodings, considered in several ways, can be reduced by using a dialect that builds in {syntactic sugar}  for regular expression patterns, and there are many tools that make constructing such syntactic dialects simple, e.g.
-Camlp4 \cite{ocaml-manual} and Sugar* \cite{erdweg2011sugarj,erdweg2013framework} (though as we will see, they come with costs of their own).
+Camlp4 \cite{OCaml-manual} and Sugar* \cite{erdweg2011sugarj,erdweg2013framework} (though as we will see, they come with costs of their own).
 %Put another way, such embeddings may not preserve the cognitive cost associated with a construct (qualitative criteria are generally used to establish this \cite{green1996usability}).
@@
 Record types provide a simple example -- they cannot be expressed by context-independent elaboration to a language with only nullary and binary product types.
-Various  languages have explored ``record-like'' primitives that go still further, supporting functional update and extension operators, width and depth coercions (sometimes implicit)%\cite{Cardelli:1984:SMI:1096.1098}
-, methods, prototypic dispatch and other ``semantic embellishments'' that cannot be expressed by context-independent elaboration to a language with only standard record types (we will detail an  example in Sec.
+Various  languages have explored ``record-like'' primitives that go still further, supporting ``functional update'' operators, width and depth coercions (sometimes implicit)%\cite{Cardelli:1984:SMI:1096.1098}
+, methods, prototypic dispatch and other such ``semantic embellishments'' that cannot be expressed by context-independent elaboration to a language with only standard record types (we will detail an  example in Sec.
 \ref{sec:metamodules}).
-OCaml primitively builds in the type structure of ``polymorphic variants'', ``open datatypes'' and  operations that use format strings like $\mathtt{sprintf}$ \cite{ocaml-manual}.
+OCaml primitively builds in the type structure of ``polymorphic variants'', ``open datatypes'' and  operations that use format strings like $\mathtt{sprintf}$ \cite{OCaml-manual}.
 ReactiveML builds in primitives for functional reactive programming \cite{mandel2005reactiveml}.
@@
 
+%\subsection{Dialects Considered Harmful}
 % express record types as syntactic sugar over the simply-typed lambda calculus with  binary product types.\footnote{Pairs can of course be expressed as syntactic sugar atop records, though one could argue that using binary products as the more primitive concept is simpler.} The static semantics need to be extended with new type and term operators.
@@
 %At best, one can hope that the compilers for the decidedlytwo languages target a common intermediate language,
-As MLj demonstrates, addressing the interoperability problem requires somehow ``combining'' the  dialects into a single language.
-In the most general setting where the dialects in question might be specified by judgements of arbitrary form, this is not a well-defined notion.
-Even if we restrict our interest to dialects specified using formalisms that do operationalize this notion of dialect combination, there is generally no guarantee that the combined dialect will conserve syntactic and semantic properties that can be established about the dialects in isolation.
+As MLj demonstrates, addressing this interoperability problem requires somehow ``combining'' the  dialects into a single language.
+However, in the most general setting where the dialects in question might be specified by judgements of arbitrary form, this is not a well-defined notion.
+Even if we restrict our interest to dialects specified using formalisms that do operationalize this notion of dialect combination, there is generally no guarantee that the combined dialect will conserve important syntactic and semantic properties that can be established about the dialects in isolation.
 %In other words, any putative ``combined language'' must formally be considered a  distinct system for which one must derive essentially all metatheorems of interest anew, guided only informally by those derived for the dialects individually.
 %There is no well-defined mechanism for constructing such a ``combined language'' in general.
-For example, consider two derived syntactic dialects specified using context-free grammars, one specifying derived syntax for finite mappings, the other specifying a similar syntax for ordered finite mappings.
-Though each can be shown to have an unambiguous concrete syntax in isolation, when their grammars are na\"ively  combined by, for example, Camlp4,  ambiguities  arise.
+For example, consider two syntactic dialects, one specifying derived syntax for finite mappings, the other specifying a similar syntax for ordered finite mappings.
+Though each dialect can be shown to have an unambiguous concrete syntax in isolation, when their grammars are na\"ively  combined by, for example, Camlp4,  ambiguities  arise.
 % would generally attempt to combine their semantics, particularly when they are specified in very different ways.
@@
 %It it clear that dialects are better rhetorical devices than practical engineering artifacts. 
-Due to this paucity of modular reasoning principles, language dialects are not practical for software development ``in the large''.
+Due to this paucity of modular reasoning principles, language dialects are not entirely practical for software development ``in the large''.
 %Large software projects and software ecosystems must pick a single language that does provide powerful modular reasoning principles and, to benefit from them, stay inside it.
 
-Dialect designers must ultimately take a less direct approach to have an impact on large-scale software development -- they must convince the designers in control of comparatively popular languages, like Ocaml and Scala, to include some suitable variant of the primitives they espouse into backwards compatible language revisions.
+%\subsection{Reducing the Need For Dialects}
+Dialect designers must instead take a less direct approach to have an impact on large-scale software development: they must convince the designers in control of comparatively popular languages, like OCaml and Scala, to include some suitable variant of the primitives they espouse into backwards compatible language revisions.
 %These decisions are increasingly influenced by community processes, e.g.
@@
 %rimitives that are only situationally useful, or that trigger aesthetic disagreements between different factions of the community, are still quite often left languishing in impractical ``toy'' dialects.
-Recalling the words of  Reynolds, which are as relevant today as they were almost half a century ago \cite{Reynolds70}:%Because there is no data about how useful a construct is in practice until it is included in a language like this, decisions about which constructs to include are often informed by little more than intuition.
+%Recalling the words of  Reynolds, which are clearly as relevant today as they were almost half a century ago \cite{Reynolds70}:%Because there is no data about how useful a construct is in practice until it is included in a language like this, decisions about which constructs to include are often informed by little more than intuition.
 %This approach is antithetical to the ideal of a truly \emph{general-purpose language} described at the beginning of this section.
 %\newpage
-\begin{quote}\textit{The recent development of programming languages suggests that the simul\-taneous achievement of simplicity 
-and generality in language design is a serious unsolved 
-problem.}\begin{flushright}--- John Reynolds (1970)\end{flushright}
-\end{quote}
+%\begin{quote}\textit{The recent development of programming languages suggests that the simul\-taneous achievement of simplicity 
+%and generality in language design is a serious unsolved 
+%problem.}\begin{flushright}--- John Reynolds (1970)\end{flushright}
+%\end{quote}
 
@@
 % that this implies.
-The other is to search for more general primitives that reduce these seemingly \emph{ad hoc} primitives to modularly composable library constructs.
+The other is to search for more general primitives that reduce the primitives found in dialects today to modularly composable library constructs.
 % so that they can be evaluated on their individual merits by programmers and used together without the possibility of conflict.
@@
 Encouragingly, primitives of this sort do occasionally arise.
-For example, a recent revision of Ocaml added support for  ``generalized algebraic data types'' (GADTs), based on research on guarded recursive datatype constructors \cite{XiCheChe03}.
-Using GADTs, Ocaml was able to move some of the \emph{ad hoc} machinery for typechecking operations that use format strings, like \texttt{sprintf}, out of the language and into a library (however, syntactic machinery remains primitively built in).
+For example, a recent revision of OCaml added support for  ``generalized algebraic data types'' (GADTs), based on research on guarded recursive datatype constructors \cite{XiCheChe03}.
+Using GADTs, OCaml was able to move some of the \emph{ad hoc} machinery for typechecking operations that use format strings, like \texttt{sprintf}, out of the language and into a library (however, syntactic machinery remains primitively built in).
 
@@
 \section{Proposed Contributions}
-Our broad aim in the work being proposed is to introduce a new functional programming language called Verse\footnote{We distinguish Verse from Wyvern, which is the language referred to in prior publications about some of the work being proposed here, because Wyvern is a group effort evolving independently in some important ways.} that takes more radical steps down the second path just described by giving library providers the ability to directly express new derived concrete syntax as well as new type structure of a variety of designs in a safe and modularly composable manner atop a minimal type-theoretic core.
-As a result,  1) Verse will be smaller than comparable languages like ML and Scala, and 2) dialect formation will be less necessary.%This eliminates the needs to build in fewer \emph{ad hoc} constructs and dialects are less frequently necessary.
+Our broad aim in the work being proposed is to introduce a new functional programming language called Verse\footnote{We distinguish Verse from Wyvern, which is the language referred to in prior publications about some of the work being proposed here, because Wyvern is a group effort evolving independently in some important ways.} that takes more radical steps down the second path just described by giving library providers the ability to directly express new derived syntax as well as new type structure of a variety of designs in a safe and modularly composable manner atop a minimal type-theoretic \emph{internal language}.
+By taking this approach, 1) Verse will be smaller than comparable languages like ML and Scala, and 2) dialect formation will be less frequently necessary.%This eliminates the needs to build in fewer \emph{ad hoc} constructs and dialects are less frequently necessary.
 %thereby reducing the need for language dialects and revisions.
@@
 \end{enumerate} 
-Both TSMs and metamodules involve \emph{static code generation} (also called \emph{static} or \emph{compile-time metaprogramming}), in that the relevant rules in the static semantics call for the evaluation of \emph{static functions} that generate static representations of expressions and types.
+Both TSMs and metamodules involve \emph{static code generation} (also called \emph{static} or \emph{compile-time metaprogramming}), meaning that the relevant rules in the static semantics call for the evaluation of \emph{static functions} that generate static representations of expressions and types.
 Library providers write these static functions in the Verse \emph{static language} (SL).
+This design has conceptual roots in earlier work on \emph{active libraries}, which similarly envisioned using compile-time computation to give library providers more control over various aspects of the language and compilation process \cite{active-libraries-thesis}.
 %Maintaining a separation between the static (or ``compile-time'') phase and the dynamic (or ``run-time'') phase is an important facet of Verse's design.
@@
 
-The key challenge in the design of these primitives will come in ensuring that they are metatheoretically well-behaved, given that they aim to decentralize control over aspects of the concrete syntax and type structure that, in other contemporary languages, are under the exclusive control of the language designer.
+The main challenge in the design of these primitives will come in ensuring that they are metatheoretically well-behaved, given that they aim to decentralize control over aspects of the concrete syntax and type structure that, in other contemporary languages, are under the exclusive control of the language designer.
 If we are not careful, many of the problems  that arise when combining language dialects, discussed earlier, could simply shift into the semantics of these primitives.\footnote{This is why languages  like Verse are often called \emph{extensible languages}, though this is somewhat of a misnomer.
-The chief characteristic of an extensible language is that it \emph{doesn't} need to be extended in situations where other languages would need to be extended.
+The defining characteristic of an extensible language is that it \emph{doesn't} need to be extended in situations where other languages would need to be extended.
 We will avoid this somewhat confusing terminology.} Our main technical contributions will be in showing how to address these problems in a principled manner.
 In particular, syntactic conflicts will be impossible by construction and the semantics will validate code statically generated by TSMs and metamodules to maintain a strong \emph{hygienic type discipline} and, most uniquely, powerful \emph{modular reasoning principles}.
-In other words, library providers will have the ability to reason about the constructs that they have defined in isolation, and clients will be able to use them safely in any program context (i.e.
-\emph{hygenically}) and in any combination, without the possibility of conflict (i.e.
-\emph{modularly}).\footnote{This is not quite true -- simple naming conflicts can arise.
+In other words, library providers will have the ability to reason about the constructs that they have defined in isolation, and clients will be able to use them safely in any program context and in any combination, without the possibility of conflict.\footnote{This is not quite true -- simple naming conflicts can arise.
 We will tacitly assume that they are being avoided extrinsically, e.g.
@@
 A functional programming language can give library providers the ability to %meta\-pro\-gram\-matic\-ally 
-express new derived concrete syntax and type structure atop a minimal type-theoretic internal language while maintaining a hygienic type discipline and modular reasoning principles.
+express new derived textual syntax and new type structure atop a minimal type-theoretic internal language while maintaining a hygienic type discipline and modular reasoning principles.
 %These  primitives are  expressive enough to subsume the need for a variety of primitives that are, or would need to be, built in to comparable contemporary languages.
@@
 Finally, Verse intentionally is not a dependently-typed language like Coq, Agda or Idris, because these languages do not maintain a phase separation between ``compile-time'' and ``run-time.'' This phase separation is useful for programming tasks (where one would like to be able to discover errors before running a program, particularly programs that may have an effect) but less so for theorem proving tasks (where it is mainly the fact that a pure expression is well-typed that is of interest, by the propositions-as-types principle).
-Verse is designed to be used for programming tasks where SML, Ocaml, Haskell or Scala would be used today, not for advanced theorem proving tasks.
+Verse is designed to be used for programming tasks where SML, OCaml, Haskell or Scala would be used today, not for advanced theorem proving tasks.
 That said, we conjecture that the primitives we describe could be added to languages like Gallina (the ``external language'' of the Coq proof assistant  \cite{Coq:manual}) or to the ``program extraction'' mechanisms of proof assistants like Coq with  modifications, but do not plan to pursue this line of research in this dissertation.
@@
 \section{Verse}\label{sec:verse}
-To situate ourselves within a formal framework, let us begin with a brief overview of how Verse is organized and specified. %Verse consists of a \emph{module language} atop a \emph{core language}.
+To situate ourselves within a formal framework, let us begin with a brief overview of how Verse is organized and specified at a high level.
+We will assume that readers are familiar with the early chapters of a textbook like \emph{TAPL} \cite{tapl} or \emph{PFPL} \cite{pfpl}.%Verse consists of a \emph{module language} atop a \emph{core language}.
 
@@
 The key novelty is that the core language is split into a user-facing \emph{typed external language} (EL) specified by type-directed translation to a minimal \emph{typed internal language} (IL).
-One might think of this design as shifting the first stage of a type-directed compiler (e.g.
+Notionally, one might think of this design as shifting the first stage of a type-directed compiler (e.g.
 the TIL compiler for Standard ML \cite{tarditi+:til-OLD})  ``one level up'' into the language itself.
-A third sublanguage, the \emph{typed static language} (SL), is involved in giving library providers programmatic control over aspects of this type-directed translation from the EL to the IL.
-This design has conceptual roots in earlier work on \emph{active libraries}, which similarly envisioned using compile-time computation to give library providers more control over various aspects of the language and compilation process \cite{active-libraries-thesis}.% The SL is notionally analagous to the language that the first stage itself is written in.
-
-Technically, we take a standard type theoretic approach (we will assume that readers are familiar with the early chapters of a textbook like \emph{TAPL} \cite{tapl} or \emph{PFPL} \cite{pfpl}).
-We summarize the main judgements relevant to the IL, EL and SL below.
-Readers who prefer to see examples first can skim these sections for now, returning to them as needed.
+A third sublanguage, the \emph{static language} (SL), is involved in giving library providers programmatic control over aspects of this type-directed translation from the EL to the IL.
+% The SL is notionally analagous to the language that the first stage itself is written in.
+We summarize the main judgements relevant to the IL, EL and SL below. Readers who prefer to see examples first can skim these sections for now, returning as needed.
 
@@
 \textbf{Judgement Form} & \textbf{Pronunciation}\\
-\vdash \iota : \tau & \text{Internal expression $\iota$ has internal type $\tau$.}\\
-\vdash \tau~\mathtt{itype} & \text{Internal type $\tau$ is valid.}
+\vdash \tau~\mathtt{itype} & \text{Internal type $\tau$ is valid.}\\
+\vdash \iota : \tau & \text{Internal expression $\iota$ is assigned internal type $\tau$.}
 \end{array}
@@
 However, our design is largely insensitive to such details -- the only strict requirements are that the IL be type safe and support parametric type abstraction.
-To give a simple account of our novel contributions, so we will stick to this small, familiar IL for the purposes of this work.
+To give a simple account of our novel contributions, we will stick to this small, familiar IL for the purposes of this work.
 %Features like record types and labeled sum types, on the other hand, are not built in.
@@
 \subsubsection{External Language}
-The external language supports a richer syntax and type structure atop the IL.
+The external language (EL) supports a richer syntax and type structure atop the IL.
 Most programmers will interact exclusively with the EL, so throughout this work, the words ``expression'' and ``type'' used without qualification refer to external expressions and types, respectively.
-The main judgements in the specification of the EL take the following form  (again omitting various contexts, which we will introduce progressively):
+The central judgements in the specification of the EL take the following form  (again omitting various contexts, which we will introduce progressively):
 \\[1ex]
@@
 \textbf{Judgement Form} & \textbf{Pronunciation}\\
+\vdash \sigma~\mathtt{type} \leadsto \tau & \text{Static expression $\sigma$ is a type with translation $\tau$.}\\
 \vdash e \Rightarrow \sigma \leadsto \iota & \text{Expression $e$ synthesizes type $\sigma$ and has translation $\iota$.}\\
-\vdash e \Leftarrow \sigma \leadsto \iota & \text{Expression $e$ analyzes against type $\sigma$ and has translation $\iota$.}\\
-\vdash \sigma~\mathtt{type} \leadsto \tau & \text{Static expression $\sigma$ is a type with translation $\tau$.}
+\vdash e \Leftarrow \sigma \leadsto \iota & \text{Expression $e$ analyzes against type $\sigma$ and has translation $\iota$.}
 \end{array}
-$\\[1ex]
-Notice that the expression typing judgements are \emph{bidirectional}, i.e.
+$\\[1ex]The main points to note here are that:
+\begin{itemize}
+\item External types are distinct from internal types (thus distinguishing this style of specification from the Harper-Stone semantics for Standard ML \cite{Harper00atype-theoretic}).
+\item The expression typing judgements are \emph{bidirectional}, i.e.
 we make a judgemental distinction between \emph{type synthesis} (the type is an ``output'') and \emph{type analysis} (the type is an ``input'') \cite{Pierce:2000:LTI:345099.345100}.
 This will  allow us to explicitly specify how Verse's \emph{local type inference} works, and in particular, how it interacts with the primitives that we aim to introduce.
-Like Scala, we intentionally do not seek to support non-local type inference.
+Like Scala, we do not seek to support non-local type inference.
 %, primarily because it is not technically compatible with the primitives we will introduce (but also because we believe that compilers for languages that support only local type inference generate clearer error messages).
+\item  The dynamic behavior of an external expression is determined directly by its translation to the IL, i.e. there is no separate transition system governing the dynamics of external expressions.
+
+Proving type safety amounts to proving that the translation of every external expression of type $\sigma$ is an internal expression of type $\tau$, where $\tau$ is the translation of $\sigma$ (and thus, by type safety of the IL, evaluation does not ``go wrong'').
+This corresponds notionally to the idea of \emph{type-preserving compilation} in type-directed compilers \cite{tarditi+:til-OLD}.
+\end{itemize}
 
 \subsubsection{Static Language}
-Finally, the static language plays an essential role in the semantics of the novel primitives that are the topic of this proposal.
-The SL is another typed lambda calculus consisting of \emph{static expressions}, $\sigma$, classified by \emph{sorts}\footnote{We use the word ``sort'' to avoid confusion with the phrase ``static type''.
+Finally, the static language plays an important role in the semantics of the novel primitives that are the topic of this proposal.
+The static language is simply another typed lambda calculus consisting of \emph{static expressions}, $\sigma$, classified by \emph{sorts}\footnote{We use the word ``sort'' to avoid confusion with the phrase ``static type''.
 Our use of the  word ``sort'' should be considered distinct from the various other uses of this word found in the literature on programming languages.}, $\uppi$, according to the following judgements (the ``static statics''):
@@
 \textbf{Judgement Form} & \textbf{Pronunciation}\\
-\vdash \sigma :: \uppi & \text{Static expression $\sigma$ has sort $\uppi$.}\\
-\vdash \uppi~\mathtt{sort} & \text{Sort $\uppi$ is valid.}
+\vdash \uppi~\mathtt{sort} & \text{Sort $\uppi$ is valid.}\\
+\vdash \sigma :: \uppi & \text{Static expression $\sigma$ has sort $\uppi$.}
 \end{array}
@@
 
-The Verse module system is taken directly from Standard ML, with which we assume a working familiarity for the purposes of this proposal \cite{harper1997programming,MacQueen:1984:MSM:800055.802036} (a related module system, e.g.
-Ocaml's, would work just as well for our purposes).
+The Verse module system is taken directly (up to small syntactic changes) from Standard ML, with which we assume a working familiarity for the purposes of this proposal \cite{harper1997programming,MacQueen:1984:MSM:800055.802036} (a related module system, e.g.
+OCaml's, would work just as well for our purposes).
 We will give examples of its use in Sec.
@@
 
-Formally, our approach will be to first specify our contributions assuming a language without an ML-style module system, then make additions to this specification to make it compatible with the presence of a module system.
+Formally, our approach will be to first specify our contributions assuming a language without an ML-style module system, then make modifications to this specification to make it compatible with the presence of a module system.
 Because it has been thoroughly studied in the literature, we will defer to prior work both here and in the dissertation for many formal details.
-In particular, we will assume that modules are being tracked by a suitable \emph{module context}, without detailing how this context is populated.
+In particular, it will suffice to assume that modules are being tracked by a suitable \emph{module context}, without detailing how this context is populated.
 %This context will be analagous to the \emph{named type context} specified in the papers just cited.% Instead, we will consider it abstractly where it interacts with the core language.%\todo{reminder that we are not going to re-specify the module system}
@@
 For example, derived list syntax is built in to many functional languages, so that instead of having to write out \lstinline{Cons(1, Cons(2, Cons(3, Nil)))}, the programmer can equivalently write \lstinline{[1, 2, 3]}.
-Many languages go beyond this, building in derived syntax associated with various other types of data, like vectors (the SML/NJ dialect of SML), arrays (Ocaml), monadic commands (Haskell), syntax trees (Scala), XML documents (Scala, Ur/Web) and SQL queries (F\#, Ur/Web).
+Many languages go beyond this, building in derived syntax associated with various other types of data, like vectors (the SML/NJ dialect of SML), arrays (OCaml), monadic commands (Haskell), syntax trees (Scala, F\#), XML documents (Scala, Ur/Web) and SQL queries (F\#, Ur/Web).
 %This is a rather \emph{ad hoc} process.% discussed previously, the usual approach is to require that the language designer build in new derived syntactic forms.
@@
 We will begin in Sec.
-\ref{sec:examples} by detailing a representative example for which such a mechanism would be useful: regular expression patterns expressed using abstract data types.
+\ref{sec:examples} by detailing another example for which such a mechanism would be useful: regular expression patterns expressed using abstract data types.
 In Sec.
@@
 \ref{sec:syntax-contributions}, we outline our proposed alternatives -- \emph{typed syntax macros} (TSMs) and the related \emph{type-specific languages} (TSLs) -- and discuss how they resolve these issues.
-We also show how TSMs are specified in terms of the judgement forms that were introduced in the previous section.
+We also give an overview of how TSMs are specified in terms of the judgement forms that were introduced in the previous section.
 We conclude in Sec.
@@
 \end{lstlisting}
- Client of any module \lstinline{P} that satisfies \lstinline{PATTERN}, written \lstinline{P : PATTERN}, manipulate patterns as values of the type \verb|P.t| using the interface described by this signature.
+ Client of any module \lstinline{P} that has been opaquely ascribed \lstinline{PATTERN}, written \lstinline{P :> PATTERN}, manipulate patterns as values of the type \verb|P.t| using the interface described by this signature.
 The identity of this type is held abstract outside the module during typechecking (i.e.
-it acts as a newly generated type) so that the burden of proving that there is no way to use the case analysis function to distinguish patterns with the same normal form is local to the module, and implementation details do not escape (and can thus evolve freely).
+it acts as a newly generated type).
+As a result, the burden of proving that there is no way to use the case analysis function to distinguish patterns with the same normal form is local to the module, and implementation details do not escape (and can thus evolve freely).
 %The details are standard and not particularly relevant for our purposes, so we omit them here.
@@
 When compiling an expression like this, the client would see an error message like \verb|error: illegal escape character|.\footnote{This is the error message that \texttt{javac} produces.
-When compiling an analagous expression using SML of New Jersey (SML/NJ), we encounter the rather bizarre error message \texttt{Error: unclosed string}.} In a small lab study, we observed that this class of error was common, and often temporarily confused even experienced programmers if they had not used regular expressions recently \cite{Omar:2012:ACC:2337223.2337324}.
-The standard workaround has higher syntactic cost -- one must double all backslashes:
+When compiling an analagous expression using SML of New Jersey (SML/NJ), we encounter the rather bizarre error message \texttt{Error: unclosed string}.} In a small lab study, we observed that this class of error was common, and often confused even experienced programmers if they had not used regular expressions recently \cite{Omar:2012:ACC:2337223.2337324}.
+The standard workaround has higher syntactic cost -- we must double all backslashes:
 \begin{lstlisting}[numbers=none]
@@
 Some languages, anticipating such modes of use, build in alternative string forms that leave escape sequences uninterpreted.
-For example, Ocaml supports the following, which has only a constant syntactic cost:
+For example, OCaml supports the following, which has only a constant syntactic cost:
 \begin{lstlisting}[numbers=none]
@@
 
-Statically verifying that pattern formation errors will not dynamically arise requires reasoning about the full dynamics of the language.
-This is an undecidable verification problem in general and is difficult to even partially automate.
+Statically verifying that pattern formation errors will not dynamically arise requires reasoning about arbitrary dynamic behavior.
+This is an undecidable verification problem in general and can be difficult to even partially automate.
 In this example, the verification procedure would first need to be able to establish that the variable \lstinline{pattern} is equal to the parse function \lstinline{PU.parse}.
-If the string argument had not been written canonically but rather computed, e.g.
+If the string argument had not been written literally but rather computed, e.g.
 as \lstinline{"SSTR(GESTR" ^ "SSTRCESTR"} where \lstinline{^} is the string concatenation function applied in infix style, it would also need to be able to establish that this expression is equivalent to the string \lstinline{"SSTR(GCESTR"}.
@@
 
-\item The next problem is that dynamic string parsing mainly decreases the syntactic cost of ``literal'' patterns.
+\item The next problem is that dynamic string parsing mainly decreases the syntactic cost of complete patterns.
 Patterns constructed compositionally cannot easily benefit from this technique.
@@
 
-\item Finally, for functions like \lstinline{example} where we are constructing patterns on the basis of data of type \lstinline{string}, using strings coincidentally to introduce patterns makes it easy for programmers to use string concatenation in subtly incorrect ways.
+\item Finally, for functions like \lstinline{example} where we are constructing patterns on the basis of data of type \lstinline{string}, using strings coincidentally to introduce patterns tempt programmers to use string concatenation in subtly incorrect ways.
 For example, consider the following seemingly more readable definition of \lstinline{example}:
@@
 
-Syntax extension systems based on context-free grammars like  Sugar* \cite{erdweg2013framework}, Camlp4 \cite{ocaml-manual} and several others, as well as mechanisms that allow direct programmatic control over parsing like Racket's reader extensions \cite{Flatt:2012:CLR:2063176.2063195} are more expressive, and would allow us to directly introduce pattern syntax into our core language's grammar, perhaps following Unix conventions like this:
+Syntax extension systems that are based on context-free grammars like  Sugar* \cite{erdweg2013framework}, Camlp4 \cite{OCaml-manual} and several others would allow us to directly introduce pattern syntax into our core language's grammar, perhaps following Unix conventions like this:
 \begin{lstlisting}[numbers=none]
@@
 So properly considered, every combination of extensions introduced using these mechanisms creates a \emph{de facto} derived syntactic dialect of our language.
+The benefit of these systems over monolithic systems like Racket's syntactic preprocessor \cite{Flatt:2012:CLR:2063176.2063195} is only that they lower the implementation cost of constructing syntactic dialects.
 % Resolving such parsing amibiguities is left to each client of the library.
@@
 Here, \lstinline{pattern} is a macro parameterized by a module \lstinline{P : PATTERN}.
-It statically parses the provided string literal (which must still be written using an Ocaml-style literal here) to generate an elaboration.
+It statically parses the provided string literal (which must still be written using an OCaml-style literal here) to generate an elaboration.
 The macro specifies that the elaboration it statically generates will be of type \lstinline{P.t}.
@@
 
-Having to creatively repurpose existing syntax like this limits the impact a library provider can have on syntactic cost (particularly when there are pre-existing conventions that are not consistent with the conventions adopted by the language).
+Having to creatively repurpose existing syntax in this way limits the impact a library provider can have on syntactic cost (particularly when it would be desirable to adopt conventions that are not consistent with the conventions adopted by the language).
 It also can create confusion for readers expecting parenthesized expressions to behave in a consistent manner.
-However,  this approach is still common because it avoids many of the problems of na\"ive syntax extensions: there cannot be syntactic conflicts (because the syntax is not extended at all), we can define macro abbreviations because macros are integrated into the language, there is a hygiene discipline that guarantees that the elaboration will not capture variables inadvertently, and using a typed macro system, there is a clear type discipline (i.e.
+However,  this approach is still preferably to direct syntax extension because it avoids many of the problems discussed above: there cannot be syntactic conflicts (because the syntax is not extended at all), we can define macro abbreviations because macros are integrated into the language, there is a hygiene discipline that guarantees that the elaboration will not capture variables inadvertently, and using a typed macro system, there is a clear type discipline (i.e.
 programmers need not examine the elaboration to know what type a macro application must have).
@@
 
-Syntax defined by library providers using these primitives comes at syntactic cost comparable to that of derived syntax built in primitively by the language designer or  using a na\"ive syntax extension tool like Camlp4, without causing dialect formation.
+Syntax defined by library providers using TSMs and TSLs comes at syntactic cost comparable to that of derived syntax built in primitively by the language designer or  using a na\"ive syntax extension tool like Camlp4, without causing dialect formation.
 As such, these \emph{ad hoc} approaches can, in large part, be discarded.
@@
 Instead, they can be expressed using \emph{metamodules} (which are distinct from, though closely related to, {modules}, as we will discuss below).
-Only polymorphic function types are built primitively into the Verse EL.
+Only  functions are built primitively into the Verse EL.
 
@@
 
-Even if we were to automate this (using, for example, \emph{type macros} as  in Scala \cite{ScalaMacros2013}, adapted to a language with ML-style modules), another issue is that we can only reasonably express the introductory and projection operators by specifying their semantics as functions like this.
+Even if we were to automate this (using, for example, \emph{type macros} as  in Scala \cite{ScalaMacros2013}, adapted to support Verse's modules), another issue is that we can only reasonably express the introductory and projection operators by specifying their semantics as functions like this.
 To enumerate all possible operators that arise from the opcon \lstinline{ltuple-} using functions would require constructing an encoding of every labeled tuple type that might arise from dropping one or more components from the labeled tuple in question.
@@
 For example, we might consider capturing the regular string invariant using refinements of the \lstinline{string} type.
-Clients supply refinement specifications using structured comments or annotations (which can then be processed by external refinement checkers, like SML CIDRE \cite{Davies:1997:RTC}):
+Refinement annotations can then be processed by external refinement checkers, like SML CIDRE \cite{davies1997refinement}.
+In Verse, we might supply refinement specifications using structured comments:
 \begin{lstlisting}[numbers=none]
@@
 We plan to detail these in the thesis.
-On the other hand, constructs like ML-style exceptions, Ocaml-style open datatypes or Java-style class-based object systems would not be possible because they require statically tracking which constructors/classes are in scope.
+On the other hand, constructs like ML-style exceptions, OCaml-style open datatypes or Java-style class-based object systems would not be possible because they require statically tracking which constructors/classes are in scope.
 We plan to discuss adding support for new user-defined static contexts as an avenue for future work in the dissertation.
@@
 For example, F\# \cite{syme2012expert} introduces functional datatypes and the other constructs of the ML core language, defining them in terms of constructs available in the Common Intermediate Language (CIL), a class-based object-oriented language.
-Ocaml includes a class-based object system that can be defined by type-directed translation to Caml, a language building in little more than recursive product and sum types.
+OCaml includes a class-based object system that can be defined by type-directed translation to Caml, a language building in little more than recursive product and sum types.
 
@@
 % records (identified up to reordering), 
-% records with width and depth coercions \cite{Cardelli:1984:SMI:1096.1098} and records with functional update operators (also called \emph{extensible records}) \cite{ocaml-manual}.
+% records with width and depth coercions \cite{Cardelli:1984:SMI:1096.1098} and records with functional update operators (also called \emph{extensible records}) \cite{OCaml-manual}.
 {The Haskell wiki notes that ''extensible records  are not implemented in GHC.
@@
 [...] As a result, nothing much happens.'' \cite{GHCFAQ}}%,
-%  %mutable fields \cite{ocaml-manual}, 
+%  %mutable fields \cite{OCaml-manual}, 
 % %field delegation \cite{atlang-gpce14} \todo{gpce submission} 
@@
 % % %In some cases, desirable features must be omitted entirely due to concerns about interoperability.
-F\#, for example, aimed to retain source compatibility with Ocaml code, but due to the need for bidirectional interoperability with CLI languages, it does not support features like polymorphic variants, modules or functors \cite{ocaml-manual} because they have no apparent analogs in the type system of the CLI.
+F\#, for example, aimed to retain source compatibility with OCaml code, but due to the need for bidirectional interoperability with CLI languages, it does not support features like polymorphic variants, modules or functors \cite{OCaml-manual} because they have no apparent analogs in the type system of the CLI.
 % % %\end{itemize}

### 2015-07-07 verse-icfp15-src/verse-icfp15-src.tex
Verse ICFP15 SRC submission

%\documentclass[12pt]{article}
\documentclass[preprint]{sigplanconf}
% The following \documentclass options may be useful:
%
% 10pt          To set in 10-point type instead of 9-point.
% 11pt          To set in 11-point type instead of 9-point.
% authoryear    To obtain author/year citation style instead of numeric.
\usepackage{amsmath}
\usepackage{amssymb} 
\usepackage{amsthm}
\usepackage{ stmaryrd }
\usepackage{mathpartir}
\usepackage[usenames,dvipsnames,svgnames,table]{xcolor}

\newcommand{\Verse}{{\textsf{\small Verse}}}

\definecolor{mygray}{rgb}{0.5,0.5,0.5}
\usepackage{times}
\renewcommand{\ttdefault}{txtt}
\usepackage{alltt}
\usepackage{listings}

\lstset{tabsize=2, 
basicstyle=\ttfamily\fontsize{8pt}{1em}\selectfont, 
commentstyle=\itshape\ttfamily\color{gray}, 
stringstyle=\ttfamily\color{red},
numbers=left, numberstyle=\scriptsize\color{gray}\ttfamily, language=ML,moredelim=[il][\sffamily]{?},mathescape=true,showspaces=false,showstringspaces=false,xleftmargin=15pt, morekeywords=[1]{tyfam,opfam,let,fn,val,def,casetype,objtype,metadata,of,*,datatype,new,toast,syntax,module,where,import,for,ana,syn,opcon,tycon,metasignature,metamodule,metasig,metamod,static,at,tycase,mod},deletekeywords={double,in},classoffset=0,belowskip=\smallskipamount,
moredelim=**[is][\color{red}]{SSTR}{ESTR},
moredelim=**[is][\color{OliveGreen}]{SHTML}{EHTML},
moredelim=**[is][\color{purple}]{SCSS}{ECSS},
moredelim=**[is][\color{brown}]{SSQL}{ESQL},
moredelim=**[is][\color{orange}]{SCOLOR}{ECOLOR},
moredelim=**[is][\color{magenta}]{SPCT}{EPCT}, 
moredelim=**[is][\color{gray}]{SNAT}{ENAT}, 
moredelim=**[is][\color{blue}]{SURL}{EURL},
moredelim=**[is][\color{SeaGreen}]{SQT}{EQT},
moredelim=**[is][\color{Periwinkle}]{SGRM}{EGRM},
moredelim=**[is][\color{YellowGreen}]{SID}{EID},
moredelim=**[is][\color{Sepia}]{SUS}{EUS},
deletestring=[d]{"}
}
\lstloadlanguages{Java,VBScript,XML,HTML}
\let\li\lstinline

\usepackage{url}
\usepackage{todonotes}
\lefthyphenmin=5
\sloppy

\newcommand{\moutput}{^{\color{gray}-}}
\newcommand{\rulename}[1]{(#1)}
\def \TirNameStyle #1{\small\rulename{#1}}
\renewcommand{\MathparLineskip}{\lineskiplimit=.3\baselineskip\lineskip=.3\baselineskip plus .2\baselineskip}

\newtheorem{theorem}{Theorem}
\newtheorem{lemma}{Lemma}
\newtheorem{corollary}{Corollary}
\newtheorem{definition}{Definition}
\newtheorem{tyconinvariant}{Tycon Invariant}
\newenvironment{proof-sketch}{\noindent{\emph{Proof Sketch.}}}{\qed}
\makeatletter

% Heather-added packages for the fancy table
\usepackage{longtable}
\usepackage{booktabs}
\usepackage{pdflscape}
\usepackage{colortbl}%
\newcommand{\myrowcolour}{\rowcolor[gray]{0.925}}
\usepackage{wasysym}

\renewcommand\topfraction{0.85}
\renewcommand\bottomfraction{0.85}
\renewcommand\textfraction{0.1}
\renewcommand\floatpagefraction{0.85}

\AtBeginDocument{%
 \abovedisplayskip=4pt
 \abovedisplayshortskip=4pt
 \belowdisplayskip=4pt
 \belowdisplayshortskip=4pt
}

\usepackage[colorlinks=true,allcolors=blue,breaklinks,draft=false]{hyperref}   % hyperlinks, including DOIs and URLs in bibliography
% known bug: http://tex.stackexchange.com/questions/1522/pdfendlink-ended-up-in-different-nesting-level-than-pdfstartlink

\newcommand{\liv}[1]{\lstinline{#1}}
\usepackage{enumitem}

\begin{document}

\conferenceinfo{-}{-} 
\copyrightyear{-} 
\copyrightdata{[to be supplied]}

%\titlebanner{{\tt \textcolor{Red}{{\small Under Review -- distribute within CMU only.}}}}        % These are ignored unless
%\preprintfooter{Distribute within CMU only.}   % 'preprint' option specified.

\title{Modular Syntax}
\subtitle{ (ICFP 2015 Student Research Competition)}

\authorinfo{Cyrus Omar}
           {Carnegie Mellon University}
           {comar@cs.cmu.edu}

\maketitle
\begin{abstract}
% 
% This has not been a systematic process: metatheoretic results must be reestablished for each combination of fragments, guided only informally by metatheorems derived for simpler systems.
% As the language design space grows, mechanisms that systematize this process and provide stronger modular reasoning principles than this are needed.

Practical functional programming languages like ML and Haskell often include derived syntactic forms that capture common idioms more concisely or naturally. For example, derived list syntax is built into all major functional languages. Many languages, and dialects thereof, go beyond this, building in derived syntax associated with other types of data, like vectors (SML/NJ), arrays (OCaml), monadic commands (Haskell), syntax trees (Scala, F\#), XML (Scala, Ur/Web) and SQL queries (F\#, Ur/Web).

We introduce a new functional language based on ML called Verse that takes a less \emph{ad hoc} approach -- rather than privileging particular library constructs with primitive syntactic support, Verse supports primitives that allow library providers to introduce new syntactic expansions on their own, in a safe, hygienic and modular manner.\end{abstract}

%\category{D.3.2}{Programming Languages}{Language Classifications}[Extensible Languages]
%\category{D.3.4}{Programming Languages}{Processors}[Compilers]
%\category{F.3.1}{Logics \& Meanings of Programs}{Specifying and Verifying and Reasoning about Programs}[Specification Techniques]
%\keywords
%extensible languages; module systems; type abstraction; typed compilation; type-level computation

\section{Introduction}\label{intro}
Let us begin from the perspective of a regular expression library provider, where the abstract syntax of {patterns}, $p$, over strings, $s$, is specified as below:\[p ::= \textbf{empty} ~|~ \textbf{str}(s) ~|~ \textbf{seq}(p; p) ~|~ \textbf{or}(p; p) ~|~ \textbf{star}(p) ~|~ \textbf{group}(p)\]

In a language with an ML-style module system \cite{MacQueen:1984:MSM:800055.802036}, one standard way to express this abstract syntax is by defining an \emph{abstract datatype} with a signature like that shown in Figure \ref{fig:PATTERN}. Clients of any module \lstinline{P} that has been opaquely ascribed this signature, written \lstinline{P :> PATTERN}, manipulate patterns as values of the type \verb|P.t| using the interface described by this signature. The identity of this type is held abstract outside the module during typechecking (i.e. it acts as a newly generated type). This is beneficial in that it ensures that implementation details (e.g. internal compiled representations of patterns) do not escape (and thus the library provider can reason about them modularly and evolve them freely).

\begin{figure}
\begin{lstlisting}[deletekeywords={case},numbers=none]
signature PATTERN = sig {
  type t
  val Empty : t
  val Str : string -> t
  val Seq : t * t -> t
  val Or : t * t -> t
  val Star : t -> t
  val Group : t -> t
  val case : (
    t -> 
    'a -> (* Empty case *)
    (string -> 'a) -> (* Str case *)
    (t * t -> 'a) -> (* Seq case *)
    (t * t -> 'a) -> (* Or case *)
    (t -> 'a) -> (* Star case *)
    (t -> 'a) -> (* Group case *)
    'a)
}
\end{lstlisting}
\caption{A signature describing an abstract datatype.}
\label{fig:PATTERN}
\end{figure}
 %An example is discussed in the supplement.%We make only syntactic changes, discussed in the supplement, consistent with Verse's layout-sensitive concrete syntax.

The abstract syntax of patterns is too verbose to be practical  in all but the most trivial examples, so programmers conventionally write patterns using a more concise concrete syntax. For example, the concrete syntax \lstinline{A|T|G|C} corresponds to the following much more verbose pattern expression:
\begin{lstlisting}[numbers=none,mathescape=|]
P.Or(P.Str "SSTRAESTR", P.Or(P.Str "SSTRTESTR", 
	P.Or(P.Str "SSTRGESTR", P.Str "SSTRCESTR")))
\end{lstlisting}

\section{Existing Approaches}\label{sec:syntax-existing}
\subsection{Dynamic String Parsing}
To expose this more concise concrete syntax for patterns to clients, the most common approach is to provide a function that parses strings to produce patterns. Because, as just mentioned, there  may be many implementations of the \lstinline{PATTERN} signature, the usual approach is to define a parameterized module (a.k.a. a \emph{functor} in SML) defining utility functions like this abstractly:

\begin{lstlisting}[numbers=none]
module PatternUtil(P : PATTERN) => mod {
  fun parse(s : string) : P.t => (* ... *)
}
\end{lstlisting}
This allows a client of any module \lstinline{P : PATTERN} 
to construct patterns like this:
\begin{lstlisting}[numbers=none]
let module PU = PatternUtil(P)
PU.parse "SSTRA|T|G|CESTR"
\end{lstlisting}
 %Again, none of our goals are comprehensively achieved.
Unfortunately, this approach is imperfect for several reasons:
\begin{enumerate} 
\item The syntax of strings conflicts with the syntax for patterns (e.g. pattern escape sequences like \li{\d} are misparsed as invalid string escape sequences). Doubling backslashes (and other workarounds) increases syntactic cost.
\item Pattern parsing does not occur until a pattern is evaluated, so malformed patterns are not detected statically. This can also incur run-time cost, which must be managed carefully.
\item Idioms for constructing patterns compositionally are not captured by this technique (we will see an example below).
\item Using strings coincidentally to introduce patterns tempts programmers to use string concatenation in subtly incorrect ways when working with string data. For example, consider the following function:
\begin{lstlisting}[numbers=none,escapechar=|]
fun example_bad(name : string) => 
  pattern (name ^ "SSTR: \\d\\d\\d\\dESTR")
\end{lstlisting}

This function works correctly for common inputs (i.e. alphabetic names), but when the input contains special characters that have meaning in the concrete syntax of patterns, a problem arises. In applications that query sensitive data, mistakes like this lead to \emph{injection attacks}, which are among the most common and catastrophic security threads on the web today.
\end{enumerate}

\subsection{Direct Syntax Extension}\label{sec:syntax-extension}
One tempting alternative to dynamic string parsing is to use a system that gives the users of a language the power to directly extend its concrete syntax with new derived forms like  Sugar* \cite{erdweg2013framework} or Camlp4 \cite{OCaml-manual}. These allow us to directly introduce pattern syntax into our core language's grammar, perhaps following Unix conventions like this:
\begin{lstlisting}[numbers=none]
let val ssn = /\d\d\d-\d\d-\d\d\d\d/
\end{lstlisting}

For patterns constructed compositionally, we can define \emph{splicing syntax} that allows us to write string and pattern expressions inline (distinguished by prefixes \lstinline{@} and \lstinline{%}, respectively). 
For example:
\begin{lstlisting}[numbers=none,escapechar=|]
fun example_concise(name : string) => 
  /@name: %ssn/
\end{lstlisting}
%The body of this function elaborates to the body of \lstinline{example_fixed} as shown above. 
Had we mistakenly written \lstinline{%name}, we would encounter only a static type error, rather than the  silent injection  vulnerability discussed above.

The problem is that we cannot guarantee that \emph{syntactic conflicts} between such extensions will not arise. If another library provider used overlapping syntax for a different implementation or variant of regular expressions, or for some other unrelated construct, then a client could not simultaneously use both libraries at the same time.

\subsection{Macros}\label{sec:term-rewriting}
An alternative approach is to leave the concrete syntax of the language fixed, but repurpose it for novel ends using a \emph{macro system}. The LISP macro system \cite{Hart63a} is perhaps the most prominent example of such a system. %Early variants of this system suffered from the problem of unhygienic variable capture just described, but  later variants, notably in the Scheme dialect of LISP, brought support for enforcing hygiene \cite{Kohlbecker86a}. 
In languages with a richer static type structure, variants of macros that restrict rewriting to a particular type and perform the rewriting statically have also been studied \cite{Herman10:Theory,ganz2001macros} and integrated into languages, e.g. Scala \cite{ScalaMacros2013}.

For example, we might repurpose the infix operators that are normally used for other purposes to support string and pattern splicing as follows:

\begin{lstlisting}[numbers=none,escapechar=|]
fun example_with_macro(name : string) => 
  pattern P (name ^ "SSTR: ESTR" + ssn)
\end{lstlisting}
Here, \lstinline{pattern} is a macro parameterized by a module \lstinline{P : PATTERN}. Having to creatively repurpose existing syntax in this way limits the impact a library provider can have on syntactic cost (particularly when it would be desirable to adopt conventions that are not consistent with the conventions adopted by the language). It also can create confusion for readers expecting parenthesized expressions to behave in a consistent manner.

\section{Contributions}\label{sec:syntax-contributions}
Verse provides a new external primitive -- the \textbf{typed syntax macro} (TSM) -- that combines the syntactic flexibility of syntax extensions with the reasoning guarantees of typed macros. TSMs can be parameterized by modules, so they can be used to define syntax valid at any type defined by a module satisfying a specified signature. This addresses all of the problems brought up above.

%A typed syntax macro is invoked by applying it to a \emph{delimited form}, which can contain  arbitrary syntax in its \emph{body}.  
To introduce TSMs, let us consider the following concrete external expression:
\begin{lstlisting}[numbers=none]
pattern P /SURLA|T|G|CEURL/
\end{lstlisting}
Here, we apply a \emph{parameterized TSM}, \lstinline{pattern}, first to a module parameter, \lstinline{P}, then to a \emph{delimited form}, \lstinline{/SURLA|T|G|CEURL/}. Note that a number of alternative delimiters are also provided by Verse's concrete syntax and could equivalently be used. The TSM statically parses the \emph{body} of the provided delimited form, i.e. the characters between the delimiters (shown here in blue), and computes an \emph{expansion}, i.e. another external expression. In this case, the expansion is the expression shown in Section 1.

The definition of \lstinline{pattern} looks like this:
\begin{lstlisting}[numbers=none]
syntax pattern(P : PATTERN) at P.t {
  static fn (ps : ParseStream) : Exp => 
    (* pattern parser here *)
}
\end{lstlisting}
This TSM definition first identifies the TSM as \lstinline{pattern}, then specifies a module parameter, \lstinline{P}, which must match the signature \lstinline{PATTERN}. Note that identifying the module parameter as \lstinline{P} here is an entirely local choice, i.e. the TSM can be applied to \emph{any} module parameter matching  \lstinline{PATTERN}. This parameter is used in the type annotation \lstinline{at P.t}, which specifies that all elaborations that arise from the application of this TSM to a module \lstinline{P} and a delimited form must necessarily be of type \lstinline{P.t}. These elaborations arise by the static action of the parse function defined next, within braces. It is written as a static function of type \lstinline{ParseStream -> Exp}. The type \lstinline{ParseStream} will give the function access to the {body} of the delimited form (in blue above) and the sort \lstinline{Exp}  encodes the abstract syntax of external expressions. %So the parse function parses the body of the delimited form to generate an encoding of the elaboration.

To support splicing syntax as described in Sec. \ref{sec:syntax-extension}, the parse function must be able to extract external subexpressions directly from the parse stream. For example, consider the client code below:
\begin{lstlisting}[numbers=none]
(* TSMs can be partially applied and abbreviated *)
let syntax pat = pattern P
let val ssn = pat /SURL\d\d\d-\d\d-\d\d\d\dEURL/
fun example_tsm(name: string) => 
  pat /SURL@EURLnameSURL: %EURLssn/
\end{lstlisting}
The subexpressions \lstinline{name} and \lstinline{ssn} on the last line occur directly in the parse stream. When the parse function encounters these, it asks the \lstinline{ParseStream} to extract these as \emph{spliced expressions} for use in the elaboration. For example, the elaboration generated for the body of \lstinline{example_tsm} above would, if written concretely with spliced expressions marked, be:
\begin{lstlisting}[numbers=none]
P.Seq(P.Str(<spliced>(name)), 
	P.Seq(P.Str ": ", <spliced>(ssn)))
\end{lstlisting}
The hygiene mechanism then ensures that only these marked portions of the generated elaboration can refer to the variables at the use site, preventing inadvertent variable capture by the elaboration.

A formal specification of this mechanism is under development, but space limitations prevent us from introducing it here. In earlier work, we described a simpler variant of TSMs which do not support integration with an ML-style module system \cite{sac15}. Module parameters represent the novel contribution of this submission.
\section{Conclusion}
Typed syntax macros allow library providers to define new syntactic elaborations, while guaranteeing to clients that these can be combined arbitrarily. They support a hygienic type discipline by the use of type annotations, and integration with Verse's ML-style module system by the use of module parameters. Taken together, we believe TSMs will decrease the need for syntactic dialects and supplant the use of \emph{ad hoc} tools like Camlp4.
%\acks
%The author is grateful to Jonathan Aldrich, Robert Bocchino and anonymous referees for their useful suggestions. This work was funded by the DOE Computational Science Graduate Fellowship under grant number DE-FG02-97ER25308.
%Acknowledgments, if needed.

% We recommend abbrvnat bibliography style.

\bibliographystyle{abbrvnat}

% The bibliography should be embedded for final submission.

\bibliography{../research}
%\softraggedright
%P. Q. Smith, and X. Y. Jones. ...reference text...

\end{document}

### 2015-07-11 thesis-proposal/problem.tex
revised secs 1 and 2

@@
 Instead, its users are able to modularly  express a wide variety of useful constructs in libraries, using the comparatively small set of primitives that the language builds in.
-Functional programming languages like Standard ML (SML), OCaml and Haskell descend from mathematically elegant typed lambda calculi, diverging mainly in that they consider various pragmatic issues in addition to  essential metatheoretic issues in their design.
-For example, they build in record types, generalizing the nullary and binary product types that suffice in simpler calculi, because explicitly labeled components can be cognitively useful to human programmers.
-Similarly, they build in derived syntax (or colloquially, ``syntactic sugar'') for selected library constructs, like lists.% concision and readability.
+Functional programming language designers have long turned to minimal typed lambda calculi to come to a principled understanding of  fundamental metatheoretic issues, like type safety.
+To then design a ``full-scale''\footnote{Throughout this work, colloquialisms that should be read as having an intuitive meaning, rather than a strict mathematical meaning, will be first introduced with quotation marks.}  language, they must also consider various human factors.
+For example, major functional programming languages like Standard ML (SML) \cite{mthm97-for-dart,harper1997programming}, OCaml \cite{ocaml-manual} and Haskell \cite{jones2003haskell} all primitively build in record types, generalizing the nullary and binary product types that suffice in simpler calculi, because explicitly labeled components are cognitively useful to human programmers.
+Similarly, they build in derived syntax (or colloquially, ``syntactic sugar'') that decreases the syntactic cost of  common idioms, like list construction.% concision and readability.
+%
 %Minimal typed lambda calculi play a central role in the study of programming languages, and serve as direct inspiration for  functional programming languages like ML.
@@
 %syntactic and semantic primitives that capture common idioms naturally and precisely.
-e
-The hope is that a limited number of primitive ``embellishments'' like these will suffice to turn a minimal lambda calculus into a ``general-purpose'' programming language.
+
  %The community around this language should, ideally, then be able to direct its efforts almost exclusively toward the development of useful libraries, rather than further modifications to the language itself.
-Unfortunately, a stable language design that fully achieves this ideal has yet to emerge, as evidenced by the diverse array of \emph{dialects} that continue to proliferate around all major contemporary languages.
+One might hope that a limited number of primitive ``embellishments'' like these will suffice to go from minimal calculi to a broadly useful (or ``general-purpose'') programming language.
+But a stable language design that fully achieves this ideal has yet to emerge, as evidenced by the diverse array of ``dialects'' that continue to proliferate around all major contemporary languages.
 % For example, let us consider the many dialects of ML.
@@
 %Tools for  constructing new ``domain-specific'' dialects also continue to proliferate. % The situation is similar across all major language lineages. 
-Indeed, tools that assist with the construction of so-called  ``domain-specific'' language (DSL) dialects\footnote{Such dialects are also sometimes called ``external DSLs''.
-The terms ``internal'' and ``embedded DSL'' refer informally to programming interfaces that  ``resemble'' distinct dialects, but that are expressed as libraries, using existing primitives \cite{fowler2010domain}.} are becoming increasingly prominent.
+Indeed, tools that aid in the construction of so-called  ``domain-specific'' language dialects (DSLs)\footnote{Such dialects are also sometimes called ``external DSLs'', to distinguish them from  ``internal'' or ``embedded DSLs'', which are actually  library interfaces that  ``resemble'' distinct dialects \cite{fowler2010domain}.} are becoming increasingly prominent.
 {This calls for an investigation}: why is it that programmers and researchers are still so often unable to satisfyingly express the constructs that they need in libraries, as modes of use of the ``general-purpose'' primitives already available in major languages today?
@@
 %Library implementations are not yet satisfying.
-%\subsection{Why are there so many dialects?}
+\subsection{Why are there so many dialects?}
 Perhaps the most common reason for this ongoing proliferation of dialects may simply be that the \emph{syntactic cost} of expressing a construct of interest using contemporary general-purpose primitives is not always ideal.
 To decrease syntactic cost, library providers often construct \emph{syntactic dialects}, i.e.
-dialects that can be specified by a purely syntactic (i.e.
+dialects that can be specified by purely syntactic (i.e.
 context-independent) elaboration to the existing language.
+Put another way, they introduce only new derived syntactic forms.
 For example, Ur/Web is a syntactic dialect of Ur (a language that itself descends from ML) that builds in derived syntax for SQL queries, HTML elements and other datatypes used in the domain of web programming \cite{conf/popl/Chlipala15}.
@@
 \ref{sec:syntax}.
-We will be able to express precisely the semantics that we seek, but existing approaches for approximating the concrete syntax of patterns, e.g.
+%We will be able to express precisely the semantics that we seek, but existing approaches for approximating the concrete syntax of patterns, e.g.
 using dynamic string parsing,  will leave something to be desired.
@@
 derived list syntax in SML \cite{harper1997programming,mthm97-for-dart}.
-Tools like Camlp4 \cite{OCaml-manual}, Sugar* \cite{erdweg2011sugarj,erdweg2013framework} and Racket \cite{Flatt:2012:CLR:2063176.2063195} have lowered the engineering costs of constructing syntactic dialects in situations like these, contributing to their ongoing proliferation.
+Tools like Camlp4 \cite{ocaml-manual}, Sugar* \cite{erdweg2011sugarj,erdweg2013framework} and Racket \cite{Flatt:2012:CLR:2063176.2063195} have lowered the engineering costs of constructing syntactic dialects in such situations, contributing to their proliferation.
 %(we will consider them further in Sec.
@@
 More advanced dialects introduce new type structure, going beyond what is possible with derived syntax.
-Record types provide a simple example -- they cannot be expressed by context-independent elaboration to a language with only nullary and binary product types.
+As a simple example, the static and dynamic semantics of records cannot be expressed by context-independent elaboration to a language with only nullary and binary products.
 Various  languages have explored ``record-like'' primitives that go still further, supporting ``functional update'' operators, width and depth coercions (sometimes implicit)%\cite{Cardelli:1984:SMI:1096.1098}
-, methods, prototypic dispatch and other such ``semantic embellishments'' that cannot be expressed by context-independent elaboration to a language with only standard record types (we will detail an  example in Sec.
+, methods, prototypic dispatch and other such ``semantic embellishments'' that cannot be expressed by context-independent expansion to a language with only standard record types (we will detail an  example in Sec.
 \ref{sec:metamodules}).
-OCaml primitively builds in the type structure of ``polymorphic variants'', ``open datatypes'' and  operations that use format strings like $\mathtt{sprintf}$ \cite{OCaml-manual}.
+OCaml primitively builds in the type structure of polymorphic variants, open datatypes and  operations that use format strings like $\mathtt{sprintf}$ \cite{ocaml-manual}.
 ReactiveML builds in primitives for functional reactive programming \cite{mandel2005reactiveml}.
 ML5 builds in high-level primitives for distributed programming based on a modal lambda calculus \cite{Murphy:2007:TDP:1793574.1793585}.
-Manticore \cite{conf/popl/FluetRRSX07} and AliceML  \cite{AliceLookingGlass} build in parallel programming primitives with a more elaborate type structure than is found in core calculi for parallel programming.
+Manticore \cite{conf/popl/FluetRRSX07} and AliceML  \cite{AliceLookingGlass} build in parallel programming primitives with a more elaborate type structure than is found in simpler accounts of parallelism.
 MLj builds in the type structure of the Java object system (motivated by a desire to interface safely and naturally with Java libraries) \cite{Benton:1999:IWW:317636.317791}.
@@
 
-%\subsection{Dialects Considered Harmful}
+\subsection{Dialects Considered Harmful}
 % express record types as syntactic sugar over the simply-typed lambda calculus with  binary product types.\footnote{Pairs can of course be expressed as syntactic sugar atop records, though one could argue that using binary products as the more primitive concept is simpler.} The static semantics need to be extended with new type and term operators.
@@
 However, in the most general setting where the dialects in question might be specified by judgements of arbitrary form, this is not a well-defined notion.
-Even if we restrict our interest to dialects specified using formalisms that do operationalize this notion of dialect combination, there is generally no guarantee that the combined dialect will conserve important syntactic and semantic properties that can be established about the dialects in isolation.
+Even if we restrict our interest to dialects specified using formalisms that do operationalize the notion of dialect combination, there is generally no guarantee that the combined dialect will conserve important syntactic and semantic properties that can be established about the dialects in isolation.
 %In other words, any putative ``combined language'' must formally be considered a  distinct system for which one must derive essentially all metatheorems of interest anew, guided only informally by those derived for the dialects individually.
@@
 %It it clear that dialects are better rhetorical devices than practical engineering artifacts. 
-Due to this paucity of modular reasoning principles, language dialects are not entirely practical for software development ``in the large''.
+Due to this paucity of modular reasoning principles, the ``dialect-oriented'' approach (also called the ``language-oriented approach'' \cite{journals/stp/Ward94}) is not sensible for software development ``in the large''.
 %Large software projects and software ecosystems must pick a single language that does provide powerful modular reasoning principles and, to benefit from them, stay inside it.
@@
 \section{Proposed Contributions}
-Our broad aim in the work being proposed is to introduce a new functional programming language called Verse\footnote{We distinguish Verse from Wyvern, which is the language referred to in prior publications about some of the work being proposed here, because Wyvern is a group effort evolving independently in some important ways.} that takes more radical steps down the second path just described by giving library providers the ability to directly express new derived syntax as well as new type structure of a variety of designs in a safe and modularly composable manner atop a minimal type-theoretic \emph{internal language}.
-By taking this approach, 1) Verse will be smaller than comparable languages like ML and Scala, and 2) dialect formation will be less frequently necessary.%This eliminates the needs to build in fewer \emph{ad hoc} constructs and dialects are less frequently necessary.
+Our broad aim in the work being proposed is to introduce new primitive constructs that give library providers the ability to directly express new syntactic expansions as well as new type structure of a variety of designs in a safe and modularly composable manner atop a minimal type-theoretic \emph{internal language}.
+As a vehicle for this work, we will describe a new functional programming language called Verse.\footnote{We distinguish Verse from Wyvern, which is the language referred to in prior publications about some of the work being proposed here, because Wyvern is a group effort evolving independently in some important ways.} We formally introduce the semantics of Verse in Sec.
+\ref{sec:verse}.
+By supporting the primitives that we introduce, 1) Verse will be smaller than comparable languages like ML and Scala, and 2) dialect formation will be less frequently necessary.
+In other words, these primitives reduce the need for many others:%This eliminates the needs to build in fewer \emph{ad hoc} constructs and dialects are less frequently necessary.
 %thereby reducing the need for language dialects and revisions.
@@
 Unlike SML, the Verse core language is split into a \emph{typed external language} (EL) specified by {type-directed translation} to a minimal \emph{typed internal language} (IL).
-We formally introduce the semantics of Verse in Sec. \ref{sec:verse}. Our novel contributions will be two primitives that eliminate the need for many others:
 \begin{enumerate}
 \item \textbf{Typed syntax macros} (TSMs), introduced in Sec.
-\ref{sec:syntax}, eliminate the need to primitively build in derived concrete syntax specific to library constructs, e.g.
-list syntax as in SML or XML syntax as in Scala and Ur/Web, by giving library providers static control over the parsing and elaboration of delimited segments of concrete syntax (at a specified type or parameterized family of types).
+\ref{sec:syntax}, reduce the need to primitively build in derived concrete syntax specific to library constructs, e.g.
+list syntax as in SML or XML syntax as in Scala and Ur/Web, by giving library providers static control over the parsing and expansion of delimited segments of concrete syntax (at a specified type or parameterized family of types).
 %We show how TSMs can be invoked explicitly by clients, or, implicitly based on local type inference.
 \item \textbf{Metamodules}, introduced in Sec.
-\ref{sec:metamodules}, eliminate the need to primitively build in the type structure of constructs like records (and variants thereof),  labeled sums and other more esoteric constructs that we will introduce later by giving library providers programmatic hooks directly into the semantics.
+\ref{sec:metamodules}, reduce the need to primitively build in the type structure of constructs like records (and variants thereof),  labeled sums and other more esoteric constructs that we will introduce later by giving library providers programmatic hooks directly into the semantics.
 %For example, a library provider can implement the type structure of records with a metamodule that:
@@
 Both TSMs and metamodules involve \emph{static code generation} (also called \emph{static} or \emph{compile-time metaprogramming}), meaning that the relevant rules in the static semantics call for the evaluation of \emph{static functions} that generate static representations of expressions and types.
-Library providers write these static functions in the Verse \emph{static language} (SL).
-This design has conceptual roots in earlier work on \emph{active libraries}, which similarly envisioned using compile-time computation to give library providers more control over various aspects of the language and compilation process \cite{active-libraries-thesis}.
+Library providers write these static functions using the Verse \emph{static language} (SL).
+This design has conceptual roots in earlier work on \emph{active libraries}, which similarly envisioned using compile-time computation to give library providers more control over aspects of the language and compilation process \cite{active-libraries-thesis}.
 %Maintaining a separation between the static (or ``compile-time'') phase and the dynamic (or ``run-time'') phase is an important facet of Verse's design.
@@
 A functional programming language can give library providers the ability to %meta\-pro\-gram\-matic\-ally 
-express new derived textual syntax and new type structure atop a minimal type-theoretic internal language while maintaining a hygienic type discipline and modular reasoning principles.
+express new syntactic expansions and new type structure atop a minimal type-theoretic internal language while maintaining a hygienic type discipline and modular reasoning principles.
 %These  primitives are  expressive enough to subsume the need for a variety of primitives that are, or would need to be, built in to comparable contemporary languages.
@@
 
-It may also be useful to explicitly acknowledge that library providers could leverage the primitives we introduce   to define constructs that are in rather poor taste.
+It may also be useful to explicitly acknowledge that library providers could leverage the primitives we introduce   to define constructs that are in ``poor taste''.
 We  expect that in practice, Verse will come with a standard library defining a carefully curated collection of standard constructs, as well as guidelines for advanced users regarding when it would be sensible to use the mechanisms we introduce (following the example of languages that support operator overloading or type classes \cite{Hall:1996:TCH:227699.227700}, which also have the potential for such ``abuse'').
@@
 
-Verse takes a less \emph{ad hoc} approach -- rather than privileging  particular library constructs with primitive syntactic support, Verse exposes primitives that allow library providers to introduce new elaboration logic on their own, in a safe and modular manner.
+Verse takes a less \emph{ad hoc} approach -- rather than privileging  particular library constructs with primitive syntactic support, Verse exposes primitives that allow library providers to introduce new expansion logic on their own, in a safe and modular manner.
 %Lists need no special consideration from the language specification.
@@
 Let us begin by taking the perspective of a regular expression library provider.
-We assume the reader has some familiarity with regular expressions \cite{Thompson:1968:PTR:363347.363387}.
+We assume the reader has some familiarity with regular expressions \cite{Thompson:1968:PTR:363347.363387}\todo{better ref}.
 We will discuss a standard variant of regular expressions that supports marking \emph{captured groups} (with parentheses in the concrete syntax) to make certain examples more interesting.
@@
 \end{lstlisting}
- Client of any module \lstinline{P} that has been opaquely ascribed \lstinline{PATTERN}, written \lstinline{P :> PATTERN}, manipulate patterns as values of the type \verb|P.t| using the interface described by this signature.
+ Clients of any module \lstinline{P} that has been opaquely ascribed \lstinline{PATTERN}, written \lstinline{P :> PATTERN}, manipulate patterns as values of the type \verb|P.t| using the interface described by this signature.
 The identity of this type is held abstract outside the module during typechecking (i.e.
@@
 \begin{lstlisting}[numbers=none]
-module P : PATTERN = mod {
+module P :> PATTERN = mod {
   type t = (* ... *)

### 2015-07-12 thesis-proposal/problem.tex
revised secs 1 and 2 further

@@
 Perhaps the most common reason for this ongoing proliferation of dialects may simply be that the \emph{syntactic cost} of expressing a construct of interest using contemporary general-purpose primitives is not always ideal.
-To decrease syntactic cost, library providers often construct \emph{syntactic dialects}, i.e.
+In response, library providers construct \emph{syntactic dialects}, i.e.
 dialects that can be specified by purely syntactic (i.e.
 context-independent) elaboration to the existing language.
-Put another way, they introduce only new derived syntactic forms.
+In other words, they introduce only new derived forms.
 For example, Ur/Web is a syntactic dialect of Ur (a language that itself descends from ML) that builds in derived syntax for SQL queries, HTML elements and other datatypes used in the domain of web programming \cite{conf/popl/Chlipala15}.
@@
 \section{Proposed Contributions}
-Our broad aim in the work being proposed is to introduce new primitive constructs that give library providers the ability to directly express new syntactic expansions as well as new type structure of a variety of designs in a safe and modularly composable manner atop a minimal type-theoretic \emph{internal language}.
-As a vehicle for this work, we will describe a new functional programming language called Verse.\footnote{We distinguish Verse from Wyvern, which is the language referred to in prior publications about some of the work being proposed here, because Wyvern is a group effort evolving independently in some important ways.} We formally introduce the semantics of Verse in Sec.
-\ref{sec:verse}.
-By supporting the primitives that we introduce, 1) Verse will be smaller than comparable languages like ML and Scala, and 2) dialect formation will be less frequently necessary.
+%Our broad aim in the work being proposed is to introduce primitive language mechanisms that give library providers the ability to  express new syntactic expansions as well as new types and operators in a safe and modularly composable manner.
+Our aim in the work being proposed is to introduce primitive language constructs that take further steps down the second path just described.
+In particular, we plan to introduce the following constructs:
+% By supporting the primitives that we introduce, 1) Verse will be smaller than comparable languages like ML and Scala, and 2) dialect formation will be less frequently necessary.
 In other words, these primitives reduce the need for many others:%This eliminates the needs to build in fewer \emph{ad hoc} constructs and dialects are less frequently necessary.
@@
 %\end{enumerate}
-We will see direct analogies between ML-style modules (which Verse also supports) and metamodules in Sec. \ref{sec:metamodules}.
+We will see direct analogies between ML-style modules and metamodules in Sec. \ref{sec:metamodules}.
 \end{enumerate} 
-Both TSMs and metamodules involve \emph{static code generation} (also called \emph{static} or \emph{compile-time metaprogramming}), meaning that the relevant rules in the static semantics call for the evaluation of \emph{static functions} that generate static representations of expressions and types.
-Library providers write these static functions using the Verse \emph{static language} (SL).
+Both TSMs and metamodules involve \emph{static code generation} (also called \emph{static} or \emph{compile-time metaprogramming}), meaning that the relevant rules in the static semantics of the language call for the evaluation of \emph{static functions} that generate static representations of expressions and types.
+%Library providers write these static functions using the Verse \emph{static language} (SL).
 This design has conceptual roots in earlier work on \emph{active libraries}, which similarly envisioned using compile-time computation to give library providers more control over aspects of the language and compilation process \cite{active-libraries-thesis}.
@@
 
-The main challenge in the design of these primitives will come in ensuring that they are metatheoretically well-behaved, given that they aim to decentralize control over aspects of the concrete syntax and type structure that, in other contemporary languages, are under the exclusive control of the language designer.
+The main challenge in the design of these primitives will come in ensuring that they are metatheoretically well-behaved.
 If we are not careful, many of the problems  that arise when combining language dialects, discussed earlier, could simply shift into the semantics of these primitives.\footnote{This is why languages  like Verse are often called \emph{extensible languages}, though this is somewhat of a misnomer.
 The defining characteristic of an extensible language is that it \emph{doesn't} need to be extended in situations where other languages would need to be extended.
-We will avoid this somewhat confusing terminology.} Our main technical contributions will be in showing how to address these problems in a principled manner.
+We will avoid this somewhat confusing terminology.} Our main technical contributions will be in rigorously showing how to address these problems in a principled manner.
 In particular, syntactic conflicts will be impossible by construction and the semantics will validate code statically generated by TSMs and metamodules to maintain a strong \emph{hygienic type discipline} and, most uniquely, powerful \emph{modular reasoning principles}.
@@
 by using a URI-based naming scheme as in the Java ecosystem.} We will make these notions completely precise as we continue.
+
+As a vehicle for this work, we will describe a new functional programming language called Verse and specify  minimal calculi.\footnote{We distinguish Verse from Wyvern, which is the language referred to in prior publications about some of the work being proposed here, because Wyvern is a group effort evolving independently in some important ways.} We will give a brief overview of Verse as well as these calculi in Sec.
+\ref{sec:verse}.
 
@@
 Verse is designed to be used for programming tasks where SML, OCaml, Haskell or Scala would be used today, not for advanced theorem proving tasks.
-That said, we conjecture that the primitives we describe could be added to languages like Gallina (the ``external language'' of the Coq proof assistant  \cite{Coq:manual}) or to the ``program extraction'' mechanisms of proof assistants like Coq with  modifications, but do not plan to pursue this line of research in this dissertation.
+That said, we conjecture that the primitives we describe could be added to languages like Gallina (the ``external language'' of the Coq proof assistant  \cite{Coq:manual}) or to the program extraction mechanisms of proof assistants like Coq with  modifications, but do not plan to pursue this line of research in this dissertation.
 %Our interest is in making sure that the standard library is no more privileged than any other library.

### 2015-07-12 thesis-proposal/problem.tex
Revised Sec 3

@@
 
-As a vehicle for this work, we will describe a new functional programming language called Verse and specify  minimal calculi.\footnote{We distinguish Verse from Wyvern, which is the language referred to in prior publications about some of the work being proposed here, because Wyvern is a group effort evolving independently in some important ways.} We will give a brief overview of Verse as well as these calculi in Sec.
-\ref{sec:verse}.
+As vehicles for this work, we plan to formally specify typed lambda calculi that capture each of the novel primitives that we introduce ``minimally''.
+We will also describe (but not formally specify) a new ``full-scale'' functional language called Verse.\footnote{We distinguish Verse from Wyvern, which is the language referred to in prior publications about some of the work being proposed here, because Wyvern is a group effort evolving independently in some important ways.} The reason we will not follow Standard ML \cite{mthm97-for-dart} in giving a complete formal specification of Verse is both to emphasize that the primitives we introduce can be considered for inclusion in a variety of language designs, and to avoid distracting the reader with specifications for ``orthogonal'' primitives that are already well-understood in the literature.
+%We anticipate that future full-scale language specifications will be able to combine the ideas  in the proposed work without trouble.
+%The purpose of the work being proposed is to serve as a reference for those interested in the new constructs we introduce, not to serve as a language specification.
+We will give a brief overview covering how these languages are organized in Sec. \ref{sec:verse}.
 
@@
 
-\section{Verse}\label{sec:verse}
-To situate ourselves within a formal framework, let us begin with a brief overview of how Verse is organized and specified at a high level.
+\section{Language Overview}\label{sec:verse}
+Let us begin with a brief overview of how Verse and the minimal calculi that we will develop are organized and specified at a high level.
 We will assume that readers are familiar with the early chapters of a textbook like \emph{TAPL} \cite{tapl} or \emph{PFPL} \cite{pfpl}.%Verse consists of a \emph{module language} atop a \emph{core language}.
@@
 \subsection{Core Language}
-The Verse core language -- the language of types and expressions -- will be the focus of our formal efforts.
+The Verse core language -- the language of types and expressions -- will be the focus of our efforts.
+Figure \ref{fig:overview} gives a diagrammatic overview of how the core language is organized.
 The key novelty is that the core language is split into a user-facing \emph{typed external language} (EL) specified by type-directed translation to a minimal \emph{typed internal language} (IL).
@@
 the TIL compiler for Standard ML \cite{tarditi+:til-OLD})  ``one level up'' into the language itself.
-A third sublanguage, the \emph{static language} (SL), is involved in giving library providers programmatic control over aspects of this type-directed translation from the EL to the IL.
+A third sublanguage, the \emph{static language} (SL), is involved in giving library providers programmatic control over aspects of both macro expansion (which involves generating external terms) and this type-directed translation from the EL to the IL (which involves generating internal terms).
 % The SL is notionally analagous to the language that the first stage itself is written in.
 We summarize the main judgements relevant to the IL, EL and SL below. Readers who prefer to see examples first can skim these sections for now, returning as needed.
+
+\begin{figure}
+\includegraphics[scale=0.5,clip=true,trim=0 120 0 120]{overview.pdf}\vspace{-10px}
+\caption{A diagrammatic overview of the core language.}
+\vspace{5px}
+\label{fig:overview}
+\end{figure}
 
@@
 \begin{itemize}
-\item External types are distinct from internal types (thus distinguishing this style of specification from the Harper-Stone semantics for Standard ML \cite{Harper00atype-theoretic}).
+\item External types are distinct from internal types (thus distinguishing this style of specification from the Harper-Stone elaboration semantics for Standard ML \cite{Harper00atype-theoretic}).
 \item The expression typing judgements are \emph{bidirectional}, i.e.
@@
 As suggested by the form of the external type translation judgement above, external types are static expressions.
-More specifically, external types are static values of a primitive sort written $\mathsf{Ty}$.
+More specifically, external types are static values of a primitive sort $\mathsf{Ty}$.
 We will return to how this works in Sec.

### 2015-07-14 thesis-proposal/problem.tex
updated through sec 4

@@
 To begin, let us assume a standard, ML-like semantics for both the EL and SL and consider the EL's concrete syntax.
-Verse, like most major contemporary languages, specifies a textual concrete syntax.\footnote{Although Wyvern specified a layout-sensitive concrete syntax, to avoid unnecessary distractions, we will specify a more conventional layout-insensitive concrete syntax for Verse.} %We have chosen to specify a layout-sensitive textual concrete syntax (i.e.
+Verse, like most major contemporary languages, specifies a textual concrete syntax.\footnote{Although Wyvern specified a layout-sensitive concrete syntax, to avoid unnecessary distractions, we will describe a more conventional layout-insensitive concrete syntax for Verse.} %We have chosen to specify a layout-sensitive textual concrete syntax (i.e.
 newlines and indentation are not ignored).
@@
 \end{lstlisting}
-To avoid incurring the parsing cost for each element of \lstinline{exmpl_list}, the programmer or compiler must move the parsing step out of the closure or, barring that, ensure that an appropriately tuned memoization (i.e.
+To avoid incurring the parsing cost for each element of \lstinline{exmpl_list}, the programmer or compiler must move the parsing step out of the closure (for example, by eta-reduction in this simple example) or ensure that an appropriately tuned memoization (i.e.
 caching) strategy is being used.\footnote{Anecdotally, in scientific code using major contemporary compilers, the former optimization is not automatic.
@@
 
-\item Finally, for functions like \lstinline{example} where we are constructing patterns on the basis of data of type \lstinline{string}, using strings coincidentally to introduce patterns tempt programmers to use string concatenation in subtly incorrect ways.
+\item Finally, for functions like \lstinline{example} where we are constructing patterns on the basis of data of type \lstinline{string}, using strings coincidentally to introduce patterns tempts programmers to use string concatenation in subtly incorrect ways.
 For example, consider the following seemingly more readable definition of \lstinline{example}:
@@
 Here, \lstinline{pattern} is a macro parameterized by a module \lstinline{P : PATTERN}.
-It statically parses the provided string literal (which must still be written using an OCaml-style literal here) to generate an elaboration.
-The macro specifies that the elaboration it statically generates will be of type \lstinline{P.t}.
+It statically parses the provided string literal (which must still be written using an OCaml-style literal here) to generate an expansion.
+The macro specifies that the expansion it statically generates will be of type \lstinline{P.t}.
 By parsing the string statically, we avoid the problems of dynamic string parsing for statically-known patterns.
@@
 It also can create confusion for readers expecting parenthesized expressions to behave in a consistent manner.
-However,  this approach is still preferably to direct syntax extension because it avoids many of the problems discussed above: there cannot be syntactic conflicts (because the syntax is not extended at all), we can define macro abbreviations because macros are integrated into the language, there is a hygiene discipline that guarantees that the elaboration will not capture variables inadvertently, and using a typed macro system, there is a clear type discipline (i.e.
-programmers need not examine the elaboration to know what type a macro application must have).
+However,  this approach is still preferably to direct syntax extension because it avoids many of the problems discussed above: there cannot be syntactic conflicts (because the syntax is not extended at all), we can define macro abbreviations because macros are integrated into the language, there is a hygiene discipline that guarantees that the expansion will not capture variables inadvertently, and using a typed macro system, there is a clear type discipline (i.e.
+programmers need not examine the expansion to know what type a macro application must have).
 
@@
 The TSM statically parses the \emph{body} of the provided delimited form, i.e.
-the characters between the delimiters (shown here in blue), and computes an \emph{elaboration}, i.e.
+the characters between the delimiters (shown here in blue), and computes an \emph{expansion}, i.e.
 another external expression.
-In this case, \lstinline{pattern} generates the following elaboration (written concretely):
+In this case, \lstinline{pattern} generates the following expansion (written concretely):
 
@@
 the TSM can be applied to \emph{any} module parameter matching  \lstinline{PATTERN}.
-This parameter is used in the type annotation \lstinline{at P.t}, which specifies that all elaborations that arise from the application of this TSM to a module \lstinline{P} and a delimited form must necessarily be of type \lstinline{P.t}.
-These elaborations arise by the static action of the parse function defined next, within braces.
+This parameter is used in the type annotation \lstinline{at P.t}, which specifies that all expansions that arise from the application of this TSM to a module \lstinline{P} and a delimited form must necessarily be of type \lstinline{P.t}.
+These expansions arise by the static action of the parse function defined next, within braces.
 It is written as a static function of sort \lstinline{ParseStream -> Exp}.
@@
 The subexpressions \lstinline{name} and \lstinline{ssn} on the last line occur directly in the parse stream.
-When the parse function encounters these, it asks the \lstinline{ParseStream} to extract these as \emph{spliced expressions} for use in the elaboration.
-For example, the elaboration generated for the body of \lstinline{example_tsm} above would, if written concretely with spliced expressions marked, be:
+When the parse function encounters these, it asks the \lstinline{ParseStream} to extract these as \emph{spliced expressions} for use in the expansion.
+For example, the expansion generated for the body of \lstinline{example_tsm} above would, if written concretely with spliced expressions marked, be:
 \begin{lstlisting}[numbers=none]
@@
 \end{lstlisting}
-The hygiene mechanism then ensures that only these marked portions of the generated elaboration can refer to the variables at the use site, preventing inadvertent variable capture by the elaboration.
-For example, the following would not be a valid elaboration, because it has free variables that are not inside spliced subexpressions:
+The hygiene mechanism then ensures that only these marked portions of the generated expansion can refer to the variables at the use site, preventing inadvertent variable capture by the expansion.
+For example, the following would not be a valid expansion, because it has free variables that are not inside spliced subexpressions:
 \begin{lstlisting}[numbers=none]
@@
     \mathbf{parsestream}(b) = \sigma_\text{ps}\\
-    \sigma_\text{parser}(\sigma_\text{ps}) \Downarrow \sigma_\text{elab}\\
-    \sigma_\text{elab} \uparrow e_\text{elab}\\\\
-    \Gamma; \emptyset \vdash e_\text{elab} \Leftarrow \sigma_\text{ty} \leadsto \iota_\text{trans}
+    \sigma_\text{parser}(\sigma_\text{ps}) \Downarrow \sigma_\text{exp}\\
+    \sigma_\text{exp} \uparrow e_\text{exp}\\\\
+    \Gamma; \emptyset \vdash e_\text{exp} \Leftarrow \sigma_\text{ty} \leadsto \iota_\text{trans}
 }{
@@
 \item The second premise creates a parse stream, $\sigma_\text{ps}$, of sort $\mathsf{ParseStream}$, from the body of the delimited form.
-\item The third premise applies the parse function to the parse stream to generate a static representation of the elaboration, $\sigma_\text{elab}$ (of sort $\mathsf{Exp}$).% Note that this is ``metaprogramming'' -- a static expression in the external semantics to generate an encoding of another external term.
-\item The fourth premise decodes $\sigma_\text{elab}$, producing the elaboration itself, $e_\text{elab}$.
-\item The final premise \emph{validates} the elaboration by analyzing it against the type $\sigma_\text{ty}$ specified by the TSM.
+\item The third premise applies the parse function to the parse stream to generate a static representation of the expansion, $\sigma_\text{exp}$ (of sort $\mathsf{Exp}$).% Note that this is ``metaprogramming'' -- a static expression in the external semantics to generate an encoding of another external term.
+\item The fourth premise decodes $\sigma_\text{exp}$, producing the expansion itself, $e_\text{exp}$.
+\item The final premise \emph{validates} the expansion by analyzing it against the type $\sigma_\text{ty}$ specified by the TSM.
 The {current} typing context is ``saved'' for use when a spliced subexpression is encountered during this process by setting it as the new \emph{outer context}, as indicated.
 
-Variables in the outer context are not directly available to the elaboration, e.g. they are ignored by the (syn-var) rule:
+Variables in the outer context are not directly available to the expansion, e.g. they are ignored by the (syn-var) rule:
 \begin{mathpar}
@@
 %\todo{revise this section}The main limitation of this approach is that it 
-Though elaboration validation ensures that an incorrect parser cannot threaten type safety,  proving type safety inductively is subtle because the elaboration is not a subexpression of the expression it will replace.
-There is a ``simple'' solution to this problem: we impose the restriction that the elaboration cannot itself use TSMs or TSLs.
+Though expansion validation ensures that an incorrect parser cannot threaten type safety,  proving type safety inductively is subtle because the expansion is not a subexpression of the expression it will replace.
+There is a ``simple'' solution to this problem: we impose the restriction that the expansion cannot itself use TSMs or TSLs.
 This would be somewhat awkward in practice, so we plan to briefly discuss other more flexible solutions as well.
@@
 control flow idioms.
-In fact, this is intentional, as it makes maintaining a type discipline in the presence of TSMs substantially easier -- neither humans nor tools need to examine the elaboration to determine what type it must necessarily have.
+In fact, this is intentional, as it makes maintaining a type discipline in the presence of TSMs substantially easier -- neither humans nor tools need to examine the expansion to determine what type it must necessarily have.
 We will, however, discuss other points in the design space in the dissertation.
@@
 We described and gave a more detailed formal specification of TSMs in a recently published paper \cite{sac15}, and TSLs in a paper published last year \cite{TSLs}, both in the context of the Wyvern language.
-The mechanics of defining a TSM, statically invoking the parse function, elaboration encoding/decoding and hygienic elaboration validation have all been detailed there.
+The mechanics of defining a TSM, statically invoking the parse function, expansion encoding/decoding and hygienic expansion validation have all been detailed there.
 In the dissertation, I plan to first present the formal system closely following this presentation, updating it only to ensure that it is consistent with the description above and with the work in the next section.
@@
 \section{Conclusion}
-In summary, we have proposed a new functional programming language, Verse, that gives programmers the ability to programmatically implement new derived concrete syntax and type structure using statically evaluated functions.
+In summary, we have proposed a new functional programming language, Verse, that gives programmers the ability to programmatically implement new syntactic expansions and new type structure using statically evaluated functions.
 Consequently, Verse does not need to build in constructs that comparable languages need to (or would need to) build in, like regular expression syntax, list syntax, labeled tuples and regular strings.

### 2015-07-27 thesis/intro.tex
updated thesis template

% !TEX root = omar-thesis.tex
TEST
(This is a test).

TEST
(This is a test).

TEST
(This is a test).

TEST
(This is a test).

TEST
(This is a test).

TEST
(This is a test).

TEST
(This is a test).

TEST
(This is a test).

TEST
(This is a test).

TEST a sfdasdf 
(This is a test).

TEST
(This is a test).

TEST
(This is a test).

TEST
(This is a test).

TEST
(This is a test).

TEST
(This is a test).

TEST
(This is a test).

TEST
(This is a test).

TEST
(This is a test).

TEST
(This is a test).

TEST
(This is a test).

TEST
(This is a test).

TEST
(This is a test).

TEST
(This is a test).

TEST
(This is a test).

TEST
(This is a test).

TEST
(This is a test).

TEST
(This is a test).

TEST
(This is a test).

TEST
(This is a test).

TEST
(This is a test).

TEST
(This is a test).

HELLO

GOODSBYE

### 2015-07-31 atlang-modularity15/test.py
forgot to push mod15 updates

@@ -1,0 +1,10 @@
+from numpy import (ndarray, 
+  array)
+
+def f(x): 
+  """Put stuff here."""
+  {
+  x : int, # x is an integer
+  y : int  # y is an integer
+  } > int
+  return x

### 2015-07-31 thesis/intro.tex
Updated thesis organization

@@
 % !TEX root = omar-thesis.tex
-TEST
-(This is a test).
+\chapter{Introduction}
+\section{Motivation}
+Functional programming language designers often turn to minimal typed lambda calculi to develop a principled understanding of  fundamental metatheoretic issues, like type safety, and to examine the essential character of various primitives of interest.
+To design a ``full-scale''\footnote{Throughout this work, colloquialisms that should be read as having an intuitive meaning, rather than a strict mathematical meaning, will be first introduced with quotation marks.}  language, language designers must carefully ``combine'' such primitives and develop various generalizations and primitive ``embellishments'' motivated by human factors.
+For example, major functional programming languages like Standard ML (SML) \cite{mthm97-for-dart,harper1997programming}, OCaml \cite{ocaml-manual} and Haskell \cite{jones2003haskell} all primitively build in record types (generalizing the nullary and binary product types that suffice in simpler settings) because explicitly labeled components are cognitively useful to human programmers.
+Similarly, they all build in derived syntax (colloquially, ``syntactic sugar'') that decreases the syntactic cost of  common idioms, like list construction.% concision and readability.
 
-TEST
-(This is a test).
+%
+%Minimal typed lambda calculi play a central role in the study of programming languages, and serve as direct inspiration for  functional programming languages like ML.
+to be useful in practice, it must present a decidedly more elaborate concrete syntax and type structure to programmers, designed considering various human factors in addition to fundamental metatheoretic issues.
+For example, while it is quite common to study minimal calculi that build in only nullary and binary product types, Standard ML (SML) builds in record types, because labels are cognitively useful.
+%syntactic and semantic primitives that capture common idioms naturally and precisely.
 
-TEST
-(This is a test).
+ %The community around this language should, ideally, then be able to direct its efforts almost exclusively toward the development of useful libraries, rather than further modifications to the language itself.
+One might hope that a limited number of semantic primitives and  primitive ``embellishments'' like these will suffice to go from minimal calculi to a broadly useful (or ``general-purpose'') programming language.
+But a stable language design that fully achieves this ideal has yet to emerge, as evidenced by the diverse array of ``dialects'' that continue to proliferate around all major contemporary languages.
+% For example, let us consider the many dialects of ML.
+Though they generally agree on the importance of a powerful module system atop a core language based conceptually on the   polymorphic lambda calculus with recursive sum and product types, they all introduce various primitive syntactic and semantic extensions and tweaks of their own design (we cover several examples below).
+%Tools for  constructing new ``domain-specific'' dialects also continue to proliferate. % The situation is similar across all major language lineages. 
+Indeed, tools that aid in the construction of so-called  ``domain-specific'' language dialects (DSLs)\footnote{Such dialects are also sometimes called ``external DSLs'', to distinguish them from  ``internal'' or ``embedded DSLs'', which are actually  library interfaces that  ``resemble'' distinct dialects \cite{fowler2010domain}.} are becoming increasingly prominent.
+{This calls for an investigation}: why is it that programmers and researchers are still so often unable to satisfyingly express the constructs that they need in libraries, as modes of use of the ``general-purpose'' primitives already available in major languages today?
+%Let us investigate why this is.
+%What kinds of constructs are not expressible in languages like SML?
 
-TEST
-(This is a test).
+% Unlike libraries organized by a module system, language dialects are not trivially composable.%Constructing a software system using libraries written in different dialects is difficult because they cannot always be composed.
+%This suggests that library-based mechanisms are still not general enough to encompass many desirable modes of expression.
+%Library implementations are not yet satisfying.
+\subsection*{Why are there so many dialects?}
+Perhaps the most common reason for this ongoing proliferation of dialects may simply be that the \emph{syntactic cost} of expressing a construct of interest using contemporary general-purpose primitives is not always ideal.
+In response, library providers construct \emph{syntactic dialects}, i.e.
+dialects that can be specified by purely syntactic (i.e.
+context-independent) elaboration to the existing language.
+In other words, they introduce only new derived forms.
+For example, Ur/Web is a syntactic dialect of Ur (a language that itself descends from ML) that builds in derived syntax for SQL queries, HTML elements and other datatypes used in the domain of web programming \cite{conf/popl/Chlipala15}.
+%Syntactic cost is often assessed qualitatively \cite{green1996usability}, though quantitative metrics can be defined.
+These are far from the only types of data that could similarly benefit from the availability of specialized derived syntax.
+As another example, we will consider regular expression patterns expressed using abstract data types (as a mode of use of the ML module system) in Sec.
+\ref{sec:syntax}.
+%We will be able to express precisely the semantics that we seek, but existing approaches for approximating the concrete syntax of patterns, e.g.
+using dynamic string parsing,  will leave something to be desired.
+% An extension of Moscow ML provided syntax for monads encoded using the module system, inspired by the  syntax built in to Haskell.\todo{asked for reference, if not found will say that "Harper reports..."} %In fact, nearly all major programming languages primitively build in derived syntax for constructs that can otherwise be expressed in the standard library, e.g.
+derived list syntax in SML \cite{harper1997programming,mthm97-for-dart}.
+Tools like Camlp4 \cite{ocaml-manual}, Sugar* \cite{erdweg2011sugarj,erdweg2013framework} and Racket \cite{Flatt:2012:CLR:2063176.2063195} have lowered the engineering costs of constructing syntactic dialects in such situations, contributing to their proliferation.
+%(we will consider them further in Sec.
+\ref{sec:syntax}).
+% contributing to their proliferation.
 
-TEST
-(This is a test).
+%Many other data structures can similarly benefit from a decrease in syntactic cost.
+For example, we  will consider {regular expression patterns} encoded using abstract data types in Sec.
+\ref{sec:syntax}.
+%The syntactic cost of  such encodings, considered in several ways, can be reduced by using a dialect that builds in {syntactic sugar}  for regular expression patterns, and there are many tools that make constructing such syntactic dialects simple, e.g.
+Camlp4 \cite{OCaml-manual} and Sugar* \cite{erdweg2011sugarj,erdweg2013framework} (though as we will see, they come with costs of their own).
+%Put another way, such embeddings may not preserve the cognitive cost associated with a construct (qualitative criteria are generally used to establish this \cite{green1996usability}).
 
-TEST
-(This is a test).
+More advanced dialects introduce new type structure, going beyond what is possible with derived syntax.
+As a simple example, the static and dynamic semantics of records cannot be expressed by context-independent elaboration to a language with only nullary and binary products.
+Various  languages have explored ``record-like'' primitives that go still further, supporting ``functional update'' operators, width and depth coercions (sometimes implicit)%\cite{Cardelli:1984:SMI:1096.1098}
+, methods, prototypic dispatch and other such ``semantic embellishments'' that cannot be expressed by context-independent expansion to a language with only standard record types (we will detail an  example in Sec.
+\ref{sec:metamodules}).
+OCaml primitively builds in the type structure of polymorphic variants, open datatypes and  operations that use format strings like $\mathtt{sprintf}$ \cite{ocaml-manual}.
+ReactiveML builds in primitives for functional reactive programming \cite{mandel2005reactiveml}.
+ML5 builds in high-level primitives for distributed programming based on a modal lambda calculus \cite{Murphy:2007:TDP:1793574.1793585}.
+Manticore \cite{conf/popl/FluetRRSX07} and AliceML  \cite{AliceLookingGlass} build in parallel programming primitives with a more elaborate type structure than is found in simpler accounts of parallelism.
+MLj builds in the type structure of the Java object system (motivated by a desire to interface safely and naturally with Java libraries) \cite{Benton:1999:IWW:317636.317791}.
+Other dialects do the same for other foreign languages, e.g.
+Furr and Foster describe a dialect of OCaml that builds in the type structure of C \cite{Furr:2005:CTS:1065010.1065019}.
+Tools like proof assistants and logical frameworks are used to specify and reason metatheoretically about dialects like these, and tools like compiler generators and language frameworks \cite{erdweg2013state} lower their implementation cost, again contributing to their proliferation.
 
-TEST
-(This is a test).
+%We will also discuss labeled products, which unlike records maintain a row ordering (this makes it possible to introduce them without explicit labels, but eliminate them using labels).
 
-TEST
-(This is a test).
+\subsection*{Dialects Considered Harmful}
+% express record types as syntactic sugar over the simply-typed lambda calculus with  binary product types.\footnote{Pairs can of course be expressed as syntactic sugar atop records, though one could argue that using binary products as the more primitive concept is simpler.} The static semantics need to be extended with new type and term operators.
+However, the simplest way to express the dynamic semantics of the newly introduced term operators is by translation to nested binary products, so we can leave the operational semantics alone.
+\todo{fill this out} %For example, there are dozens of constructs that go by the name of ``records'' in various languages, each defined by a slightly different collection of primitive operations.
+\todo{examples} %, encouraged  historically  by the availability of tools like compiler generators and,  more recently, language workbenches \cite{workbenches} and DSL frameworks \cite{dsl}.
+Unfortunately, taking this approach makes it substantially more difficult for clients to import high-level abstractions orthogonally.
+% test 
+The reason why this proliferation of language dialects should be considered alarming is that it is, in an important sense, anti-modular: a library written in one dialect cannot, in general, safely and idiomatically interface with a library written in another dialect.
+%At best, one can hope that the compilers for the decidedlytwo languages target a common intermediate language,
+As MLj demonstrates, addressing this interoperability problem requires somehow ``combining'' the  dialects into a single language.
+However, in the most general setting where the dialects in question might be specified by judgements of arbitrary form, this is not a well-defined notion.
+Even if we restrict our interest to dialects specified using formalisms that do operationalize the notion of dialect combination, there is generally no guarantee that the combined dialect will conserve important syntactic and semantic properties that can be established about the dialects in isolation.
+%In other words, any putative ``combined language'' must formally be considered a  distinct system for which one must derive essentially all metatheorems of interest anew, guided only informally by those derived for the dialects individually.
+%There is no well-defined mechanism for constructing such a ``combined language'' in general.
+For example, consider two syntactic dialects, one specifying derived syntax for finite mappings, the other specifying a similar syntax for ordered finite mappings.
+Though each dialect can be shown to have an unambiguous concrete syntax in isolation, when their grammars are na\"ively  combined by, for example, Camlp4,  ambiguities  arise.
+% would generally attempt to combine their semantics, particularly when they are specified in very different ways.
+%the ideas housed in one dialect are often only available to programmers willing to do without ideas housed in other dialects.
+%It is thus infeasible to simply allow different contributors to a software system to choose their own favorite dialect for each component they are responsible for.
+%A complex software system written in multiple distinct language dialects is far too unwieldy for it to be viable.
+%It it clear that dialects are better rhetorical devices than practical engineering artifacts. 
+Due to this paucity of modular reasoning principles, the ``dialect-oriented'' approach (also called the ``language-oriented approach'' \cite{journals/stp/Ward94}) is not sensible for software development ``in the large''.
+%Large software projects and software ecosystems must pick a single language that does provide powerful modular reasoning principles and, to benefit from them, stay inside it.
 
-TEST
-(This is a test).
+%\subsection{Reducing the Need For Dialects}
+Dialect designers must instead take a less direct approach to have an impact on large-scale software development: they must convince the designers in control of comparatively popular languages, like OCaml and Scala, to include some suitable variant of the primitives they espouse into backwards compatible language revisions.
+%These decisions are increasingly influenced by community processes, e.g.
+the Scala Improvement Process.
+%This approach concentrates power as well as responsibility over maintaining metatheoretic guarantees in the hands of a small group of language designers, though increasingly influenced by various community processes (e.g.
+the Scala Improvement Process).
+%Dialects thus serve the role of rhetorical vehicles for new ideas, rather than direct artifacts. 
+%Over time, accepting such extensions has caused these languages to balloon in size. 
+This \emph{ad hoc} approach is not sustainable, for three main reasons.
+First, as suggested by the diversity of examples given above, there are simply too  many potentially useful such primitives, and many of these are only relevant in relatively narrow application domains (for derived syntax, our group has  gathered initial data speaking to this \cite{TSLs}).
+Second, primitives introduced earlier in a language's lifespan end up monopolizing finite ``syntactic resources'', forcing subsequent primitives to use ever more esoteric forms.
+And third, primitives that prove to be flawed in some way cannot be removed or changed without breaking backwards compatibility.
+%rimitives that are only situationally useful, or that trigger aesthetic disagreements between different factions of the community, are still quite often left languishing in impractical ``toy'' dialects.
+%Recalling the words of  Reynolds, which are clearly as relevant today as they were almost half a century ago \cite{Reynolds70}:%Because there is no data about how useful a construct is in practice until it is included in a language like this, decisions about which constructs to include are often informed by little more than intuition.
+%This approach is antithetical to the ideal of a truly \emph{general-purpose language} described at the beginning of this section.
+%\newpage
+%\begin{quote}\textit{The recent development of programming languages suggests that the simul\-taneous achievement of simplicity 
+%and generality in language design is a serious unsolved 
+%problem.}\begin{flushright}--- John Reynolds (1970)\end{flushright}
+%\end{quote}
 
-TEST a sfdasdf 
-(This is a test).
+%A measured approach to incorporating new primitives into a general-purpose programming language is  sensible because once a primitive is introduced, it becomes entrenched as-is and monopolizes ``syntactic resources'', as we will discuss below.
+This leaves the subset of the  language design community interested in keeping general-purpose languages small and free of \emph{ad hoc} primitives with two possible paths forward.
+One, exemplified (arguably) by SML, is to simply eschew the introduction of specialized syntax and type structure and settle on the existing primitives, which can be said to sit at a ``sweet spot'' in the overall language design space (accepting that in some circumstances, this trades away expressive power or leads to  high syntactic cost).
+% that this implies.
+The other is to search for more general primitives that reduce the primitives found in dialects today to modularly composable library constructs.
+% so that they can be evaluated on their individual merits by programmers and used together without the possibility of conflict.
+% This should render the construction of new dialects increasingly unnecessary.
+Encouragingly, primitives of this sort do occasionally arise.
+For example, a recent revision of OCaml added support for  ``generalized algebraic data types'' (GADTs), based on research on guarded recursive datatype constructors \cite{XiCheChe03}.
+Using GADTs, OCaml was able to move some of the \emph{ad hoc} machinery for typechecking operations that use format strings, like \texttt{sprintf}, out of the language and into a library (however, syntactic machinery remains primitively built in).
 
-TEST
-(This is a test).
+%Similarly, it recently introduced ``open datatypes'', which subsume its previous more specialized exception type, and captures many use cases for .
 
-TEST
-(This is a test).
+%Viewed ``dually'', one might equivalently ask for a language that builds in a core that is as small as possible, but provides expressive power comparable to languages with much larger cores.
+This is our goal in the work being proposed.
 
-TEST
-(This is a test).
+%\vspace{-10px}
+\section{Contributions}
+%Our broad aim in the work being proposed is to introduce primitive language mechanisms that give library providers the ability to  express new syntactic expansions as well as new types and operators in a safe and modularly composable manner.
+Our aim in the work being proposed is to introduce primitive language constructs that take further steps down the second path just described.
+In particular, we plan to introduce the following constructs:
+% By supporting the primitives that we introduce, 1) Verse will be smaller than comparable languages like ML and Scala, and 2) dialect formation will be less frequently necessary.
+In other words, these primitives reduce the need for many others:%This eliminates the needs to build in fewer \emph{ad hoc} constructs and dialects are less frequently necessary.
+%thereby reducing the need for language dialects and revisions.
 
-TEST
-(This is a test).
+%Verse features a module system taken directly from SML.
+Unlike SML, the Verse core language is split into a \emph{typed external language} (EL) specified by {type-directed translation} to a minimal \emph{typed internal language} (IL).
+\begin{enumerate}
+\item \textbf{Typed syntax macros} (TSMs), introduced in Sec.
+\ref{sec:syntax}, reduce the need to primitively build in derived concrete syntax specific to library constructs, e.g.
+list syntax as in SML or XML syntax as in Scala and Ur/Web, by giving library providers static control over the parsing and expansion of delimited segments of concrete syntax (at a specified type or parameterized family of types).
+%We show how TSMs can be invoked explicitly by clients, or, implicitly based on local type inference.
+\item \textbf{Metamodules}, introduced in Sec.
+\ref{sec:metamodules}, reduce the need to primitively build in the type structure of constructs like records (and variants thereof),  labeled sums and other more esoteric constructs that we will introduce later by giving library providers programmatic hooks directly into the semantics.
+%For example, a library provider can implement the type structure of records with a metamodule that:
+%\begin{enumerate}
+%\item introduces a type constructor, \lstinline{record}, parameterized by finite mappings from labels to types, and defines, programmatically, a translation to unary and binary product types (which are built in to the internal language); and
+%\item introduces operators used to work with records, minimally record introduction and elimination (but perhaps also various functional update operators), and directly implements the logic governing their typechecking and translation to the IL (which builds in only nullary and binary products).
+%\end{enumerate}
+We will see direct analogies between ML-style modules and metamodules in Sec. \ref{sec:metamodules}.
+\end{enumerate} 
+Both TSMs and metamodules involve \emph{static code generation} (also called \emph{static} or \emph{compile-time metaprogramming}), meaning that the relevant rules in the static semantics of the language call for the evaluation of \emph{static functions} that generate static representations of expressions and types.
+%Library providers write these static functions using the Verse \emph{static language} (SL).
+This design has conceptual roots in earlier work on \emph{active libraries}, which similarly envisioned using compile-time computation to give library providers more control over aspects of the language and compilation process \cite{active-libraries-thesis}.
+%Maintaining a separation between the static (or ``compile-time'') phase and the dynamic (or ``run-time'') phase is an important facet of Verse's design.
+% static code generation.
+%We will  also introduce a simple variant of each of these primitives that leverages Verse's support for local type inference to further reduce syntactic cost in certain common situations.
 
-TEST
-(This is a test).
+The main challenge in the design of these primitives will come in ensuring that they are metatheoretically well-behaved.
+If we are not careful, many of the problems  that arise when combining language dialects, discussed earlier, could simply shift into the semantics of these primitives.\footnote{This is why languages  like Verse are often called \emph{extensible languages}, though this is somewhat of a misnomer.
+The defining characteristic of an extensible language is that it \emph{doesn't} need to be extended in situations where other languages would need to be extended.
+We will avoid this somewhat confusing terminology.} Our main technical contributions will be in rigorously showing how to address these problems in a principled manner.
+In particular, syntactic conflicts will be impossible by construction and the semantics will validate code statically generated by TSMs and metamodules to maintain a strong \emph{hygienic type discipline} and, most uniquely, powerful \emph{modular reasoning principles}.
+In other words, library providers will have the ability to reason about the constructs that they have defined in isolation, and clients will be able to use them safely in any program context and in any combination, without the possibility of conflict.\footnote{This is not quite true -- simple naming conflicts can arise.
+We will tacitly assume that they are being avoided extrinsically, e.g.
+by using a URI-based naming scheme as in the Java ecosystem.} We will make these notions completely precise as we continue.
 
-TEST
-(This is a test).
+As vehicles for this work, we plan to formally specify typed lambda calculi that capture each of the novel primitives that we introduce ``minimally''.
+We will also describe (but not formally specify) a new ``full-scale'' functional language called Verse.\footnote{We distinguish Verse from Wyvern, which is the language referred to in prior publications about some of the work being proposed here, because Wyvern is a group effort evolving independently in some important ways.} The reason we will not follow Standard ML \cite{mthm97-for-dart} in giving a complete formal specification of Verse is both to emphasize that the primitives we introduce can be considered for inclusion in a variety of language designs, and to avoid distracting the reader with specifications for ``orthogonal'' primitives that are already well-understood in the literature.
+%We anticipate that future full-scale language specifications will be able to combine the ideas  in the proposed work without trouble.
+%The purpose of the work being proposed is to serve as a reference for those interested in the new constructs we introduce, not to serve as a language specification.
+We will give a brief overview covering how these languages are organized in Sec. \ref{sec:verse}.
 
-TEST
-(This is a test).
+\subsubsection*{Thesis Statement}
+In summary, we propose a thesis defending the following statement:
+\begin{quote}
+A functional programming language can give library providers the ability to %meta\-pro\-gram\-matic\-ally 
+express new syntactic expansions and new type structure atop a minimal type-theoretic internal language while maintaining a hygienic type discipline and modular reasoning principles.
+%These  primitives are  expressive enough to subsume the need for a variety of primitives that are, or would need to be, built in to comparable contemporary languages.
+\end{quote}
 
-TEST
-(This is a test).
+\subsubsection*{Disclaimers}
+Before we continue, it may be useful to explicitly acknowledge that completely eliminating the need for dialects would indeed be asking for too much: certain design decisions are fundamentally incompatible with others or require coordination across a language design.
+We aim only to decrease the need for dialects.% out a larger design space within a single language, Verse.%a subset of constructs that can be specified by a semantics of a certain ``shape'' specified by Verse (we will make this more specific later).
+%There is nothing ``universal'' about Verse.
 
-TEST
-(This is a test).
+It may also be useful to explicitly acknowledge that library providers could leverage the primitives we introduce   to define constructs that are in ``poor taste''.
+We  expect that in practice, Verse will come with a standard library defining a carefully curated collection of standard constructs, as well as guidelines for advanced users regarding when it would be sensible to use the mechanisms we introduce (following the example of languages that support operator overloading or type classes \cite{Hall:1996:TCH:227699.227700}, which also have the potential for such ``abuse'').
+For most programmers, using Verse should not be substantially different from using a language like ML or one of its dialects.%The vast majority of programmers should not use the primitives that we introduce directly.
 
-TEST
-(This is a test).
-
-TEST
-(This is a test).
-
-TEST
-(This is a test).
-
-TEST
-(This is a test).
-
-TEST
-(This is a test).
-
-TEST
-(This is a test).
-
-TEST
-(This is a test).
-
-TEST
-(This is a test).
-
-TEST
-(This is a test).
-
-TEST
-(This is a test).
-
-TEST
-(This is a test).
-
-TEST
-(This is a test).
-
-HELLO
-
-GOODSBYE
+Finally, Verse intentionally is not a dependently-typed language like Coq, Agda or Idris, because these languages do not maintain a phase separation between ``compile-time'' and ``run-time.'' This phase separation is useful for programming tasks (where one would like to be able to discover errors before running a program, particularly programs that may have an effect) but less so for theorem proving tasks (where it is mainly the fact that a pure expression is well-typed that is of interest, by the propositions-as-types principle).
+Verse is designed to be used for programming tasks where SML, OCaml, Haskell or Scala would be used today, not for advanced theorem proving tasks.
+That said, we conjecture that the primitives we describe could be added to languages like Gallina (the ``external language'' of the Coq proof assistant  \cite{Coq:manual}) or to the program extraction mechanisms of proof assistants like Coq with  modifications, but do not plan to pursue this line of research in this dissertation.
+%Our interest is in making sure that the standard library is no more privileged than any other library.
+% Our interest is not in creating a universal language, only one that is as expressive as reasonably possible.
+%only in a subset of all constructs that can be specified in a mutually orthogonal manner by a semantics of a certain prototypic ``shape''.
+We will make this more specific later.

### 2015-07-31 thesis/language-overview.tex
Updated thesis organization

% !TEX root = omar-thesis.tex

\chapter{Language Overview}
\section{External Language}
\section{Internal Language}
\section{Static Language}
\section{Module Language}

### 2015-07-31 thesis/syntax-motivation.tex
Updated thesis organization

% !TEX root = omar-thesis.tex

\chapter{Motivation}
\section{Motivating Examples}
\subsection{Lists}
\subsection{HTML}
\subsection{Regular Expressions}
\subsection{Monadic Commands}
\subsection{Quasiquotation}
\section{Existing Approaches}
\subsection{Dynamic String Parsing}
\subsection{Direct Syntax Extension}

Related work I haven't mentioned yet:
\begin{itemize}
\item Fan: http://zhanghongbo.me/fan/start.html
\item Well-Typed Islands Parse Faster: \\\url{http://www.ccs.neu.edu/home/ejs/papers/tfp12-island.pdf}
\end{itemize}
\subsection{Term Rewriting}

### 2015-08-05 thesis/intro.tex
revised introductory section

--- 2015-07-27 thesis/intro.tex
@@
 % !TEX root = omar-thesis.tex
-TEST
-(This is a test).
+\chapter{Introduction}\label{chap:intro}
+\section{Motivation}\label{sec:intro-motivation}
+Functional programming language designers often study small typed lambda calculi to develop a principled understanding of  fundamental metatheoretic issues, like type safety, and to examine the mathematical character of language primitives of interest in isolation.
+These studies inform the design of ``full-scale''\footnote{Throughout this work, phrases that should be read as having an intuitive meaning, rather than a strict mathematical meaning, will be first introduced with quotation marks.} functional languages, which combine several such primitives and also typically feature various generalizations and primitive ``embellishments''  motivated by a consideration of human factors.
+For example, major languages like Standard ML (SML) \cite{mthm97-for-dart,harper1997programming}, OCaml \cite{ocaml-manual} and Haskell \cite{jones2003haskell} all primitively build in record types, generalizing the nullary and binary product types that suffice in simpler calculi, because explicitly labeled components are cognitively useful to human programmers.
+Similarly, these languages build in derived syntactic forms (colloquially, ``syntactic sugar'') that decrease the syntactic cost of  working with common library constructs, like lists.
 
-TEST
-(This is a test).
+The hope amongst many language designers is that a limited number of primitives like these will suffice to produce a ``general-purpose'' programming language, i.e.
+one where programmers can direct their efforts almost exclusively toward the development of useful libraries.
+However, a stable language design that fully achieves this ideal has yet to emerge, as evidenced by the diverse array of ``dialects'' that continue to proliferate around all major contemporary languages.
+In fact, tools that aid in the construction of so-called  ``domain-specific'' language dialects (DSLs)\footnote{In some parts of the literature, such dialects are called ``external DSLs'', to distinguish them from  ``internal'' or ``embedded DSLs'', which are actually  library interfaces that only ``resemble'' distinct dialects \cite{fowler2010domain}.} seem only to be becoming increasingly prominent.
 
-TEST
-(This is a test).
+\subsection*{Why are there so many dialects?}
+{This calls for an investigation}: why is it that programmers and researchers are still so often unable to satisfyingly express the constructs that they need in libraries, as modes of use of the ``general-purpose'' primitives already available in major languages today, and instead see a need for new language dialects?
 
-TEST
-(This is a test).
+Perhaps the most common motivation may be that the \emph{syntactic cost} of expressing a construct of interest using contemporary general-purpose primitives is not always ideal.
+In response, library providers construct \emph{syntactic dialects} -- dialects that introduce only new derived syntactic forms.
+%Put another way, syntactic dialects can be specified by a context-independent expansion to the existing language that they are based on. 
+For example, Ur/Web is a syntactic dialect of Ur (a language that itself descends from ML \cite{conf/pldi/Chlipala10}) that builds in derived forms for SQL queries, HTML elements and other datatypes used in the domain of web programming \cite{conf/popl/Chlipala15}.
+%Syntactic cost is often assessed qualitatively \cite{green1996usability}, though quantitative metrics can be defined.
+These are certainly not the only types of data that could similarly benefit from the availability of specialized derived forms -- we will consider a number of other examples in Sec.
+\ref{sec:syntax-examples}.
+Tools like Camlp4 \cite{ocaml-manual}, Sugar* \cite{erdweg2011sugarj,erdweg2013framework} and Racket \cite{Flatt:2012:CLR:2063176.2063195}, which we will discuss in Sec.
+\ref{sec:syntax-existing-approaches}, have lowered the engineering costs of constructing syntactic dialects in such situations, contributing to their proliferation.
 
-TEST
-(This is a test).
+More advanced dialects introduce new type structure, going beyond what is possible with only new derived forms.
+As a simple example, the static and dynamic semantics of records cannot be expressed by context-independent expansion to a language with only nullary and binary products.
+Various languages have explored ``record-like'' primitives that go further, supporting functional update operators, width and depth coercions (sometimes implicit)%\cite{Cardelli:1984:SMI:1096.1098}
+, methods, prototypic dispatch and other such ``semantic embellishments'' that in turn cannot be expressed by context-independent expansion to a language with only standard record types (we will detail an  example in Sec.
+\ref{sec:metamodules-motivating-examples}).
+OCaml primitively builds in the type structure of polymorphic variants, open datatypes and  operations that use format strings like $\mathtt{sprintf}$ \cite{ocaml-manual}.
+ReactiveML builds in primitives for functional reactive programming \cite{mandel2005reactiveml}.
+ML5 builds in high-level primitives for distributed programming based on a modal lambda calculus \cite{Murphy:2007:TDP:1793574.1793585}.
+Manticore \cite{conf/popl/FluetRRSX07} and AliceML  \cite{AliceLookingGlass} build in parallel programming primitives with a more elaborate type structure than is found in simpler accounts of parallelism.
+MLj builds in the type structure of the Java object system (motivated by a desire to interface safely and naturally with Java libraries) \cite{Benton:1999:IWW:317636.317791}.
+Other dialects do the same for other foreign languages, e.g.
+Furr and Foster describe a dialect of OCaml that builds in the type structure of C \cite{Furr:2005:CTS:1065010.1065019}.
+Tools like proof assistants and logical frameworks are used to specify and reason metatheoretically about dialects like these, and tools like compiler generators and language frameworks \cite{erdweg2013state} lower their implementation cost, again contributing to their proliferation.
 
-TEST
-(This is a test).
+\subsection*{Dialects Considered Harmful}
+One might view the ongoing proliferation of dialects described above as harmless or even positive, because programmers can simply choose the right language for the job at hand \cite{journals/stp/Ward94}.
+However, this ``dialect-oriented'' approach is, in an important sense, anti-modular: a library written in one dialect cannot, in general, safely and idiomatically interface with a library written in another dialect.
+As the study of MLj demonstrated, addressing this interoperability problem requires somehow ``combining'' the  dialects into a single language.
+However, in the most general setting where the dialects in question might be specified by judgements of arbitrary form, this is not a well-defined notion.
+Even if we restrict our interest to dialects specified using formalisms that do operationalize some notion of dialect combination, there is generally no guarantee that the combined dialect will conserve important syntactic and semantic properties that can be established about the dialects in isolation.
+%In other words, any putative ``combined language'' must formally be considered a  distinct system for which one must derive essentially all metatheorems of interest anew, guided only informally by those derived for the dialects individually.
+%There is no well-defined mechanism for constructing such a ``combined language'' in general.
+For example, consider two syntactic dialects, one specifying derived syntax for finite mappings, the other specifying a similar syntax for \emph{ordered} finite mappings.
+Though each dialect can be specified by an unambiguous grammar in isolation, when these grammars are na\"ively  combined by, for example, Camlp4,  ambiguities  arise.
+%It is thus infeasible to simply allow different contributors to a software system to choose their own favorite dialect for each component they are responsible for. 
+%It it clear that dialects are better rhetorical devices than practical engineering artifacts. 
+Due to this paucity of modular reasoning principles, the ``language-oriented'' approach is problematic for software development ``in the large''.
+%Large software projects and software ecosystems must pick a single language that does provide powerful modular reasoning principles and, to benefit from them, stay inside it.
 
-TEST
-(This is a test).
+Dialect designers have instead had to take a less direct approach to have an impact on large-scale software development: they have had to convince the designers in control of comparatively popular languages, like OCaml and Scala, to include some suitable variant of the primitives they've espoused into backwards compatible language revisions.
+%These decisions are increasingly influenced by community processes, e.g.
+the Scala Improvement Process.
+%This approach concentrates power as well as responsibility over maintaining metatheoretic guarantees in the hands of a small group of language designers, though increasingly influenced by various community processes (e.g.
+the Scala Improvement Process).
+%Dialects thus serve the role of rhetorical vehicles for new ideas, rather than direct artifacts. 
+%Over time, accepting such extensions has caused these languages to balloon in size. 
+This \emph{ad hoc} approach is not sustainable, for three main reasons.
+First, as suggested by the diversity of examples given above, there are simply too  many potentially useful such primitives, and many of these are only relevant in relatively narrow application domains (for derived syntax, our group has  gathered initial data speaking to this \cite{TSLs}).
+Second, primitives introduced earlier in a language's lifespan can end up monopolizing finite ``syntactic resources'', forcing subsequent primitives to use ever more esoteric forms.
+And third, primitives that prove to be flawed in some way cannot be removed or changed without breaking backwards compatibility.
+%Because there is no data about how useful a construct is in practice until it is available in a major language, decisions about which constructs to include are often informed only by intuition.
+%Recalling the words of  Reynolds, which are clearly as relevant today as they were almost half a century ago \cite{Reynolds70}: %This approach is antithetical to the ideal of a truly \emph{general-purpose language} described at the beginning of this section.
+%\newpage
+%\begin{quote}\textit{The recent development of programming languages suggests that the simul\-taneous achievement of simplicity 
+%and generality in language design is a serious unsolved 
+%problem.}\begin{flushright}--- John Reynolds (1970)\end{flushright}
+%\end{quote}
 
-TEST
-(This is a test).
+This suggests that language designers should be quite careful and strive to keep general-purpose languages small and free of \emph{ad hoc} primitives.
+Given that we also want to avoid dialect formation, for the reasons described above, this leaves  two possible paths forward.
+One, exemplified (arguably) by SML, is to simply eschew further syntactic conveniences and innovative type structure and settle on the existing primitives, which might be considered to sit at a ``sweet spot'' in the overall language design space.
+% (accepting that in some circumstances, this trades away expressive power or leads to  high syntactic cost).
+The other path forward is to search for a small number of highly general primitives that allow us degrade many of the constructs that are built primitively into languages today insead to modularly composable library constructs.
+Encouragingly, primitives of this sort do occasionally arise.
+For example, a recent revision of OCaml added support for  ``generalized algebraic data types'' (GADTs), based on research on guarded recursive datatype constructors \cite{XiCheChe03}.
+Using GADTs, OCaml was able to move some of the \emph{ad hoc} machinery for typechecking operations that use format strings, like \texttt{sprintf}, out of the language and into a library (note, however, that syntactic machinery remains  built in).
 
-TEST
-(This is a test).
+%Similarly, it recently introduced ``open datatypes'', which subsume its previous more specialized exception type, and captures many use cases for .
 
-TEST a sfdasdf 
-(This is a test).
+%Viewed ``dually'', one might equivalently ask for a language that builds in a core that is as small as possible, but provides expressive power comparable to languages with much larger cores.
+This is our goal in the work being proposed.
 
-TEST
-(This is a test).
+%\vspace{-10px}
+\section{Contributions}\label{sec:contributions}
+%Our broad aim in the work being proposed is to introduce primitive language mechanisms that give library providers the ability to  express new syntactic expansions as well as new types and operators in a safe and modularly composable manner.
+In this work, we introduce primitive language constructs that take further steps down the second path just described. In particular, we introduce the following primitives:
 
-TEST
-(This is a test).
+\begin{enumerate}
+\item \textbf{Typed syntax macros} (TSMs), introduced in Part \ref{part:syntax}, reduce the need to primitively build in derived concrete syntactic forms specific to library constructs (e.g.
+list syntax as in SML or XML syntax as in Scala and Ur/Web), by giving library providers static control over the parsing and expansion of delimited segments of textual syntax (at a specified type or parameterized family of types).
+We begin by showing explicitly-invoked TSMs, then show how to associate a privileged TSM with a type declaration and then rely on a local type inference scheme to invoke these implicitly.
+\item \textbf{Metamodules}, introduced in Part \ref{part:metamodules}, reduce the need to primitively build in the type structure of constructs like records (and variants thereof),  labeled sums and other interesting constructs that we will introduce later by giving library providers programmatic ``hooks'' directly into the semantics, which are specified as a \emph{type-directed translation semantics} (introduced in Chap.
+\ref{chap:language-overview}).
+%For example, a library provider can implement the type structure of records with a metamodule that:
+%\begin{enumerate}
+%\item introduces a type constructor, \lstinline{record}, parameterized by finite mappings from labels to types, and defines, programmatically, a translation to unary and binary product types (which are built in to the internal language); and
+%\item introduces operators used to work with records, minimally record introduction and elimination (but perhaps also various functional update operators), and directly implements the logic governing their typechecking and translation to the IL (which builds in only nullary and binary products).
+%\end{enumerate}
+%We will see direct analogies between ML-style modules (which our mechanisms also support) and metamodules later.
+\end{enumerate} 
+Both TSMs and metamodules make extensive use of \emph{static code generation} (also called \emph{static} or \emph{compile-time metaprogramming}), meaning that the relevant rules in the static semantics of the language call for the evaluation of \emph{static functions} that generate static representations of expressions and types.
+%Library providers write these static functions using the Verse \emph{static language} (SL).
+This design also has conceptual roots in earlier work on \emph{active libraries}, which similarly envisioned using compile-time computation to give library providers more control over aspects of the language and compilation process \cite{active-libraries-thesis}.
+%Maintaining a separation between the static (or ``compile-time'') phase and the dynamic (or ``run-time'') phase is an important facet of Verse's design.
+% static code generation.
+%We will  also introduce a simple variant of each of these primitives that leverages Verse's support for local type inference to further reduce syntactic cost in certain common situations.
 
-TEST
-(This is a test).
+As vehicles for this work, we plan to formally specify small typed lambda calculi that capture each of the novel primitives that we introduce ``minimally''.
+We will also describe (but not formally specify) a new ``full-scale'' functional language called Verse.\footnote{We distinguish Verse from Wyvern, which is the language referred to in prior publications about some of the work that we will describe, because Wyvern is a group effort evolving independently in some important ways.} The reason we will not follow Standard ML \cite{mthm97-for-dart} in giving a complete formal specification of Verse is both to emphasize that the primitives we introduce can be considered for inclusion in a variety of language designs, and to avoid distracting the reader with specifications for ``orthogonal'' primitives that are already well-understood in the literature.
+%We anticipate that future full-scale language specifications will be able to combine the ideas  in the proposed work without trouble.
+%The purpose of the work being proposed is to serve as a reference for those interested in the new constructs we introduce, not to serve as a language specification.
+We will give a brief overview covering how these languages are organized in Chap. \ref{chap:language-overview}.
 
-TEST
-(This is a test).
+The main challenge in the design of these primitives will come in ensuring that they are metatheoretically well-behaved.
+If we are not careful, many of the problems  that arise when combining language dialects, discussed earlier, could simply shift into the semantics of these primitives.\footnote{This is why languages  like Verse are often called ``extensible languages'', though this is somewhat of a misnomer.
+The defining characteristic of an extensible language is that it \emph{doesn't} need to be extended in situations where other languages would need to be extended.
+We will avoid this somewhat confusing terminology.} Our main technical contributions will be in rigorously showing how to address these problems in a principled manner.
+In particular, syntactic conflicts will be impossible by construction and the semantics will validate code statically generated by TSMs and metamodules to maintain a strong \emph{hygienic type discipline} and, most uniquely, powerful \emph{modular reasoning principles}.
+In other words, library providers will have the ability to reason about the constructs that they have defined in isolation, and clients will be able to use them safely in any program context and in any combination, without the possibility of conflict.\footnote{This is not quite true -- simple naming conflicts can arise.
+We will tacitly assume that they are being avoided extrinsically, e.g.
+by using a URI-based naming scheme as in the Java ecosystem.} We will make these notions completely precise as we continue.
 
-TEST
-(This is a test).
+\subsubsection*{Thesis Statement}
+In summary, we propose a thesis defending the following statement:
+\begin{quote}
+A functional programming language can give library providers the ability to %meta\-pro\-gram\-matic\-ally 
+express new syntactic expansions and new types and operators atop a small type-theoretic internal language while maintaining a hygienic type discipline and modular reasoning principles.
+%These  primitives are  expressive enough to subsume the need for a variety of primitives that are, or would need to be, built in to comparable contemporary languages.
+\end{quote}
 
-TEST
-(This is a test).
+\section{Disclaimers}
+Before we continue, it may be useful to explicitly acknowledge that completely eliminating the need for dialects would indeed be asking for too much: certain design decisions are fundamentally incompatible with others or require coordination across a language design.
+We aim only to decrease the need for dialects.
 
-TEST
-(This is a test).
+It may also be useful to explicitly acknowledge that library providers could leverage the primitives we introduce   to define constructs that are in ``poor taste''.
+We  expect that in practice, languages like Verse will come with a standard library defining a carefully curated collection of standard constructs, as well as guidelines for advanced users regarding when it would be sensible to use the mechanisms we introduce (following the example of languages that support features like operator overloading or type classes \cite{Hall:1996:TCH:227699.227700}, which also have the potential for such ``abuse'').
+For most programmers, using Verse should not be substantially different from using a language like ML or one of its dialects.%The vast majority of programmers should not use the primitives that we introduce directly.
 
-TEST
-(This is a test).
-
-TEST
-(This is a test).
-
-TEST
-(This is a test).
-
-TEST
-(This is a test).
-
-TEST
-(This is a test).
-
-TEST
-(This is a test).
-
-TEST
-(This is a test).
-
-TEST
-(This is a test).
-
-TEST
-(This is a test).
-
-TEST
-(This is a test).
-
-TEST
-(This is a test).
-
-TEST
-(This is a test).
-
-TEST
-(This is a test).
-
-TEST
-(This is a test).
-
-HELLO
-
-GOODSBYE
+Finally, Verse is by design not a dependently-typed language like Coq, Agda or Idris, because these languages do not maintain a phase separation between ``compile-time'' and ``run-time''.
+This phase separation is useful for programming tasks (where one would like to be able to discover errors before running a program, particularly programs that may have an effect) but less so for theorem proving tasks (where it is primarily the fact that a pure expression is well-typed that is of interest, by the propositions-as-types principle).
+Verse is designed to be used for programming tasks where SML, OCaml, Haskell or Scala would be used today, not for advanced theorem proving tasks.
+That said, we conjecture that the primitives we describe could be added to languages like Gallina (the ``external language'' of the Coq proof assistant  \cite{Coq:manual}) or to the program extraction mechanisms of proof assistants like Coq with  modifications, but do not plan to pursue this line of research in this dissertation.

### 2015-08-05 thesis/language-overview.tex
revised introductory section

@@
 % !TEX root = omar-thesis.tex
-
-\chapter{Language Overview}
-\section{External Language}
-\section{Internal Language}
-\section{Static Language}
-\section{Module Language}
+\chapter{Language Overview}\label{chap:language-overview}
+\todo{integrate this point}
+An important observation is that dialects that have substantially different type structure are still often implemented by translation to a simple and stable intermediate language.
+For example, the Glasgow Haskell Compiler's intermediate language has remained stable for many years, even as the number of Haskell dialects that the compiler supports has grown.
+Similarly, the primary Scala compiler uses the Java virtual machine (JVM) bytecode language as an intermediate language, which has also evolved relatively slowly.
+\section{External Language}\label{sec:EL-overview}
+\section{Internal Language}\label{sec:IL-overview}
+\section{Static Language}\label{sec:SL-overview}
+\section{Module Language}\label{sec:ML-overview}

### 2015-08-05 thesis/language-overview.tex
revised abstract

@@
 For example, the Glasgow Haskell Compiler's intermediate language has remained stable for many years, even as the number of Haskell dialects that the compiler supports has grown.
-Similarly, the primary Scala compiler uses the Java virtual machine (JVM) bytecode language as an intermediate language, which has also evolved relatively slowly.
+Similarly, the primary Scala compiler uses the Java virtual machine (JVM) bytecode language as an intermediate language, which has also evolved relatively slowly and independently.
 \section{External Language}\label{sec:EL-overview}

### 2015-08-13 thesis-proposal/problem.tex
Minor cleanup
Pulled changes from thesis into proposal

@@
 Instead, its users are able to modularly  express a wide variety of useful constructs in libraries, using the comparatively small set of primitives that the language builds in.
-Functional programming language designers have long turned to minimal typed lambda calculi to come to a principled understanding of  fundamental metatheoretic issues, like type safety.
-To then design a ``full-scale''\footnote{Throughout this work, colloquialisms that should be read as having an intuitive meaning, rather than a strict mathematical meaning, will be first introduced with quotation marks.}  language, they must also consider various human factors.
-For example, major functional programming languages like Standard ML (SML) \cite{mthm97-for-dart,harper1997programming}, OCaml \cite{ocaml-manual} and Haskell \cite{jones2003haskell} all primitively build in record types, generalizing the nullary and binary product types that suffice in simpler calculi, because explicitly labeled components are cognitively useful to human programmers.
-Similarly, they build in derived syntax (or colloquially, ``syntactic sugar'') that decreases the syntactic cost of  common idioms, like list construction.% concision and readability.
-%
-%Minimal typed lambda calculi play a central role in the study of programming languages, and serve as direct inspiration for  functional programming languages like ML.
-to be useful in practice, it must present a decidedly more elaborate concrete syntax and type structure to programmers, designed considering various human factors in addition to fundamental metatheoretic issues.
-For example, while it is quite common to study minimal calculi that build in only nullary and binary product types, Standard ML (SML) builds in record types, because labels are cognitively useful.
-%syntactic and semantic primitives that capture common idioms naturally and precisely.
-
- %The community around this language should, ideally, then be able to direct its efforts almost exclusively toward the development of useful libraries, rather than further modifications to the language itself.
-One might hope that a limited number of primitive ``embellishments'' like these will suffice to go from minimal calculi to a broadly useful (or ``general-purpose'') programming language.
-But a stable language design that fully achieves this ideal has yet to emerge, as evidenced by the diverse array of ``dialects'' that continue to proliferate around all major contemporary languages.
-% For example, let us consider the many dialects of ML.
-Though they generally agree on the importance of a powerful module system atop a core language based conceptually on the   polymorphic lambda calculus with recursive sum and product types, they all introduce various primitive syntactic and semantic extensions and tweaks of their own design (we cover several examples below).
-%Tools for  constructing new ``domain-specific'' dialects also continue to proliferate. % The situation is similar across all major language lineages. 
-Indeed, tools that aid in the construction of so-called  ``domain-specific'' language dialects (DSLs)\footnote{Such dialects are also sometimes called ``external DSLs'', to distinguish them from  ``internal'' or ``embedded DSLs'', which are actually  library interfaces that  ``resemble'' distinct dialects \cite{fowler2010domain}.} are becoming increasingly prominent.
-{This calls for an investigation}: why is it that programmers and researchers are still so often unable to satisfyingly express the constructs that they need in libraries, as modes of use of the ``general-purpose'' primitives already available in major languages today?
-%Let us investigate why this is.
-%What kinds of constructs are not expressible in languages like SML?
-
-% Unlike libraries organized by a module system, language dialects are not trivially composable.%Constructing a software system using libraries written in different dialects is difficult because they cannot always be composed.
-%This suggests that library-based mechanisms are still not general enough to encompass many desirable modes of expression.
-%Library implementations are not yet satisfying.
+Functional programming language designers have long studied small typed lambda calculi to develop a principled understanding of  fundamental metatheoretic issues, like type safety, and to examine the mathematical character of language primitives of interest in isolation.
+These studies have then informed the design of ``full-scale''\footnote{Throughout this work, phrases that should be read as having an intuitive meaning, rather than a strict mathematical meaning, will be first introduced with quotation marks.} functional languages, which combine several such primitives and also typically feature various generalizations and primitive ``embellishments''  motivated by a consideration of human factors.
+For example, major languages like Standard ML (SML) \cite{mthm97-for-dart,harper1997programming}, OCaml \cite{ocaml-manual} and Haskell \cite{jones2003haskell} all primitively build in record types, generalizing the nullary and binary product types that suffice in simpler calculi, because explicitly labeled components are cognitively useful to human programmers.
+Similarly, these languages build in derived syntactic forms (colloquially, ``syntactic sugar'') that decrease the syntactic cost of  working with common library constructs, like lists.
+
+The hope amongst many language designers is that a limited number of primitives like these will suffice to produce a ``general-purpose'' programming language, i.e.
+one where programmers can direct their efforts almost exclusively toward the development of useful libraries.
+However, a stable language design that fully achieves this ideal has yet to emerge, as evidenced by the diverse array of ``dialects'' that continue to proliferate around all major contemporary languages.
+In fact, tools that aid in the construction of so-called  ``domain-specific'' language dialects (DSLs)\footnote{In some parts of the literature, such dialects are called ``external DSLs'', to distinguish them from  ``internal'' or ``embedded DSLs'', which are actually  library interfaces that only ``resemble'' distinct dialects \cite{fowler2010domain}.} seem only to be becoming increasingly prominent.
+
 \subsection{Why are there so many dialects?}
-Perhaps the most common reason for this ongoing proliferation of dialects may simply be that the \emph{syntactic cost} of expressing a construct of interest using contemporary general-purpose primitives is not always ideal.
-In response, library providers construct \emph{syntactic dialects}, i.e.
-dialects that can be specified by purely syntactic (i.e.
-context-independent) elaboration to the existing language.
-In other words, they introduce only new derived forms.
-For example, Ur/Web is a syntactic dialect of Ur (a language that itself descends from ML) that builds in derived syntax for SQL queries, HTML elements and other datatypes used in the domain of web programming \cite{conf/popl/Chlipala15}.
+{This calls for an investigation}: why is it that programmers and researchers are still so often unable to satisfyingly express the constructs that they need in libraries, as modes of use of the ``general-purpose'' primitives already available in major languages today, and instead see a need for new language dialects?
+
+Perhaps the most common motivation may be that the \emph{syntactic cost} of expressing a construct of interest using contemporary general-purpose primitives is not always ideal.
+In response, library providers construct \emph{syntactic dialects} -- dialects that introduce only new derived syntactic forms.
+%Put another way, syntactic dialects can be specified by a context-independent expansion to the existing language that they are based on. 
+For example, Ur/Web is a syntactic dialect of Ur (a language that itself descends from ML \cite{conf/pldi/Chlipala10}) that builds in derived forms for SQL queries, HTML elements and other datatypes used in the domain of web programming \cite{conf/popl/Chlipala15}.
 %Syntactic cost is often assessed qualitatively \cite{green1996usability}, though quantitative metrics can be defined.
-These are far from the only types of data that could similarly benefit from the availability of specialized derived syntax.
-As another example, we will consider regular expression patterns expressed using abstract data types (as a mode of use of the ML module system) in Sec.
-\ref{sec:syntax}.
-%We will be able to express precisely the semantics that we seek, but existing approaches for approximating the concrete syntax of patterns, e.g.
-using dynamic string parsing,  will leave something to be desired.
-% An extension of Moscow ML provided syntax for monads encoded using the module system, inspired by the  syntax built in to Haskell.\todo{asked for reference, if not found will say that "Harper reports..."} %In fact, nearly all major programming languages primitively build in derived syntax for constructs that can otherwise be expressed in the standard library, e.g.
-derived list syntax in SML \cite{harper1997programming,mthm97-for-dart}.
-Tools like Camlp4 \cite{ocaml-manual}, Sugar* \cite{erdweg2011sugarj,erdweg2013framework} and Racket \cite{Flatt:2012:CLR:2063176.2063195} have lowered the engineering costs of constructing syntactic dialects in such situations, contributing to their proliferation.
-%(we will consider them further in Sec.
-\ref{sec:syntax}).
-% contributing to their proliferation.
-
-%Many other data structures can similarly benefit from a decrease in syntactic cost.
-For example, we  will consider {regular expression patterns} encoded using abstract data types in Sec.
-\ref{sec:syntax}.
-%The syntactic cost of  such encodings, considered in several ways, can be reduced by using a dialect that builds in {syntactic sugar}  for regular expression patterns, and there are many tools that make constructing such syntactic dialects simple, e.g.
-Camlp4 \cite{OCaml-manual} and Sugar* \cite{erdweg2011sugarj,erdweg2013framework} (though as we will see, they come with costs of their own).
-%Put another way, such embeddings may not preserve the cognitive cost associated with a construct (qualitative criteria are generally used to establish this \cite{green1996usability}).
-
-More advanced dialects introduce new type structure, going beyond what is possible with derived syntax.
-As a simple example, the static and dynamic semantics of records cannot be expressed by context-independent elaboration to a language with only nullary and binary products.
-Various  languages have explored ``record-like'' primitives that go still further, supporting ``functional update'' operators, width and depth coercions (sometimes implicit)%\cite{Cardelli:1984:SMI:1096.1098}
-, methods, prototypic dispatch and other such ``semantic embellishments'' that cannot be expressed by context-independent expansion to a language with only standard record types (we will detail an  example in Sec.
-\ref{sec:metamodules}).
+These are certainly not the only types of data that could similarly benefit from the availability of specialized derived forms -- we will consider another example in Sec. \ref{sec:syntax-examples}. 
+Tools like Camlp4 \cite{ocaml-manual}, Sugar* \cite{erdweg2011sugarj,erdweg2013framework} and Racket \cite{Flatt:2012:CLR:2063176.2063195}, which we will discuss in Sec.
+\ref{sec:syntax-existing-approaches}, have lowered the engineering costs of constructing syntactic dialects in such situations, contributing to their proliferation.
+
+More advanced dialects introduce new type structure, going beyond what is possible with only new derived forms.
+As a simple example, the static and dynamic semantics of records cannot be expressed by context-independent expansion to a language with only nullary and binary products.
+Various languages have explored ``record-like'' primitives that go further, supporting functional update operators, width and depth coercions (sometimes implicit)%\cite{Cardelli:1984:SMI:1096.1098}
+, methods, prototypic dispatch and other such ``semantic embellishments'' that in turn cannot be expressed by context-independent expansion to a language with only standard record types (we will detail an  example in Sec.
+\ref{sec:metamodules-motivating-examples}).
 OCaml primitively builds in the type structure of polymorphic variants, open datatypes and  operations that use format strings like $\mathtt{sprintf}$ \cite{ocaml-manual}.
@@
 
-%We will also discuss labeled products, which unlike records maintain a row ordering (this makes it possible to introduce them without explicit labels, but eliminate them using labels).
-
 \subsection{Dialects Considered Harmful}
@@
 % test 
-The reason why this proliferation of language dialects should be considered alarming is that it is, in an important sense, anti-modular: a library written in one dialect cannot, in general, safely and idiomatically interface with a library written in another dialect.
-%At best, one can hope that the compilers for the decidedlytwo languages target a common intermediate language,
-As MLj demonstrates, addressing this interoperability problem requires somehow ``combining'' the  dialects into a single language.
+Some have viewed the ongoing proliferation of dialects described above as harmless or even as desirable, arguing that programmers can simply choose the right language for the job \cite{journals/stp/Ward94}.
+However, this ``dialect-oriented'' approach is, in an important sense, anti-modular: a library written in one dialect cannot, in general, safely and idiomatically interface with a library written in another dialect.
+As the study of MLj demonstrated, addressing this interoperability problem requires somehow ``combining'' the  dialects into a single language.
 However, in the most general setting where the dialects in question might be specified by judgements of arbitrary form, this is not a well-defined notion.
-Even if we restrict our interest to dialects specified using formalisms that do operationalize the notion of dialect combination, there is generally no guarantee that the combined dialect will conserve important syntactic and semantic properties that can be established about the dialects in isolation.
+Even if we restrict our interest to dialects specified using formalisms that do operationalize some notion of dialect combination, there is generally no guarantee that the combined dialect will conserve important syntactic and semantic properties that can be established about the dialects in isolation.
 %In other words, any putative ``combined language'' must formally be considered a  distinct system for which one must derive essentially all metatheorems of interest anew, guided only informally by those derived for the dialects individually.
 %There is no well-defined mechanism for constructing such a ``combined language'' in general.
-For example, consider two syntactic dialects, one specifying derived syntax for finite mappings, the other specifying a similar syntax for ordered finite mappings.
-Though each dialect can be shown to have an unambiguous concrete syntax in isolation, when their grammars are na\"ively  combined by, for example, Camlp4,  ambiguities  arise.
-% would generally attempt to combine their semantics, particularly when they are specified in very different ways.
-%the ideas housed in one dialect are often only available to programmers willing to do without ideas housed in other dialects.
+For example, consider two syntactic dialects, one specifying derived syntax for finite mappings, the other specifying a similar syntax for \emph{ordered} finite mappings.
+Though each dialect can be specified by an unambiguous grammar in isolation, when these grammars are na\"ively  combined by, for example, Camlp4,  ambiguities  arise.
 %It is thus infeasible to simply allow different contributors to a software system to choose their own favorite dialect for each component they are responsible for. 
-%A complex software system written in multiple distinct language dialects is far too unwieldy for it to be viable.
 %It it clear that dialects are better rhetorical devices than practical engineering artifacts. 
-Due to this paucity of modular reasoning principles, the ``dialect-oriented'' approach (also called the ``language-oriented approach'' \cite{journals/stp/Ward94}) is not sensible for software development ``in the large''.
+Due to this paucity of modular reasoning principles, the ``dialect-oriented'' approach is problematic for software development ``in the large''.
 %Large software projects and software ecosystems must pick a single language that does provide powerful modular reasoning principles and, to benefit from them, stay inside it.
 
-%\subsection{Reducing the Need For Dialects}
-Dialect designers must instead take a less direct approach to have an impact on large-scale software development: they must convince the designers in control of comparatively popular languages, like OCaml and Scala, to include some suitable variant of the primitives they espouse into backwards compatible language revisions.
+Dialect designers have instead had to take a less direct approach to have an impact on large-scale software development: they have had to convince the designers in control of comparatively popular languages, like OCaml and Scala, to include some suitable variant of the primitives they've espoused into backwards compatible language revisions.
 %These decisions are increasingly influenced by community processes, e.g.
@@
 First, as suggested by the diversity of examples given above, there are simply too  many potentially useful such primitives, and many of these are only relevant in relatively narrow application domains (for derived syntax, our group has  gathered initial data speaking to this \cite{TSLs}).
-Second, primitives introduced earlier in a language's lifespan end up monopolizing finite ``syntactic resources'', forcing subsequent primitives to use ever more esoteric forms.
+Second, primitives introduced earlier in a language's lifespan can end up monopolizing finite ``syntactic resources'', forcing subsequent primitives to use ever more esoteric forms.
 And third, primitives that prove to be flawed in some way cannot be removed or changed without breaking backwards compatibility.
-%rimitives that are only situationally useful, or that trigger aesthetic disagreements between different factions of the community, are still quite often left languishing in impractical ``toy'' dialects.
-%Recalling the words of  Reynolds, which are clearly as relevant today as they were almost half a century ago \cite{Reynolds70}:%Because there is no data about how useful a construct is in practice until it is included in a language like this, decisions about which constructs to include are often informed by little more than intuition.
-%This approach is antithetical to the ideal of a truly \emph{general-purpose language} described at the beginning of this section.
+%Because there is no data about how useful a construct is in practice until it is available in a major language, decisions about which constructs to include are often informed only by intuition.
+%Recalling the words of  Reynolds, which are clearly as relevant today as they were almost half a century ago \cite{Reynolds70}: %This approach is antithetical to the ideal of a truly \emph{general-purpose language} described at the beginning of this section.
 %\newpage
@@
 
-%A measured approach to incorporating new primitives into a general-purpose programming language is  sensible because once a primitive is introduced, it becomes entrenched as-is and monopolizes ``syntactic resources'', as we will discuss below.
-This leaves the subset of the  language design community interested in keeping general-purpose languages small and free of \emph{ad hoc} primitives with two possible paths forward.
-One, exemplified (arguably) by SML, is to simply eschew the introduction of specialized syntax and type structure and settle on the existing primitives, which can be said to sit at a ``sweet spot'' in the overall language design space (accepting that in some circumstances, this trades away expressive power or leads to  high syntactic cost).
-% that this implies.
-The other is to search for more general primitives that reduce the primitives found in dialects today to modularly composable library constructs.
-% so that they can be evaluated on their individual merits by programmers and used together without the possibility of conflict.
-% This should render the construction of new dialects increasingly unnecessary.
+This suggests that language designers should be quite careful and strive to keep general-purpose languages small and free of \emph{ad hoc} primitives.
+Given that we also want to avoid dialect formation, for the reasons described above, this leaves  two possible paths forward.
+One, exemplified (arguably) by SML, is to simply eschew further syntactic conveniences and innovative type structure and settle on the existing primitives, which might be considered to sit at a ``sweet spot'' in the overall language design space.
+% (accepting that in some circumstances, this trades away expressive power or leads to  high syntactic cost).
+The other path forward is to search for a small number of highly general primitives that allow us degrade many of the constructs that are built primitively into languages today instead to modularly composable library constructs.
 Encouragingly, primitives of this sort do occasionally arise.
-For example, a recent revision of OCaml added support for  ``generalized algebraic data types'' (GADTs), based on research on guarded recursive datatype constructors \cite{XiCheChe03}.
-Using GADTs, OCaml was able to move some of the \emph{ad hoc} machinery for typechecking operations that use format strings, like \texttt{sprintf}, out of the language and into a library (however, syntactic machinery remains primitively built in).
+For example, a recent revision of OCaml added support for  generalized algebraic data types (GADTs), based on research on guarded recursive datatype constructors \cite{XiCheChe03}.
+Using GADTs, OCaml was able to move some of the \emph{ad hoc} machinery for typechecking operations that use format strings, like \texttt{sprintf}, out of the language and into a library (note, however, that syntactic machinery remains  built in).
 
@@
 Our aim in the work being proposed is to introduce primitive language constructs that take further steps down the second path just described.
-In particular, we plan to introduce the following constructs:
+In particular, we plan to introduce the following primtives:
 % By supporting the primitives that we introduce, 1) Verse will be smaller than comparable languages like ML and Scala, and 2) dialect formation will be less frequently necessary.
@@
 \begin{enumerate}
-\item \textbf{Typed syntax macros} (TSMs), introduced in Sec.
-\ref{sec:syntax}, reduce the need to primitively build in derived concrete syntax specific to library constructs, e.g.
-list syntax as in SML or XML syntax as in Scala and Ur/Web, by giving library providers static control over the parsing and expansion of delimited segments of concrete syntax (at a specified type or parameterized family of types).
-%We show how TSMs can be invoked explicitly by clients, or, implicitly based on local type inference.
+\item \textbf{Typed syntax macros} (TSMs), introduced in Sec \ref{sec:syntax}, reduce the need to primitively build in derived concrete syntactic forms specific to library constructs (e.g.
+list syntax as in SML or XML syntax as in Scala and Ur/Web), by giving library providers static control over the parsing and expansion of delimited segments of textual syntax (at a specified type or parameterized family of types).
+We begin by showing explicitly-invoked TSMs, then show how to associate a privileged TSM with a type declaration and then rely on a local type inference scheme to invoke these implicitly.
 \item \textbf{Metamodules}, introduced in Sec.
-\ref{sec:metamodules}, reduce the need to primitively build in the type structure of constructs like records (and variants thereof),  labeled sums and other more esoteric constructs that we will introduce later by giving library providers programmatic hooks directly into the semantics.
+\ref{sec:metamodules}, reduce the need to primitively build in the type structure of constructs like records (and variants thereof),  labeled sums and other interesting constructs that we will introduce later by giving library providers programmatic ``hooks'' directly into the semantics, which are specified as a \emph{type-directed translation semantics} targeting a small \emph{typed internal language} (introduced in Sec.
+\ref{sec:verse}).
 %For example, a library provider can implement the type structure of records with a metamodule that:
@@
 %\end{enumerate}
-We will see direct analogies between ML-style modules and metamodules in Sec. \ref{sec:metamodules}.
+%We will see direct analogies between ML-style modules (which our mechanisms also support) and metamodules later.
 \end{enumerate} 
-Both TSMs and metamodules involve \emph{static code generation} (also called \emph{static} or \emph{compile-time metaprogramming}), meaning that the relevant rules in the static semantics of the language call for the evaluation of \emph{static functions} that generate static representations of expressions and types.
+Both TSMs and metamodules make extensive use of \emph{static code generation} (also called \emph{static} or \emph{compile-time metaprogramming}), meaning that the relevant rules in the static semantics of the language call for the evaluation of \emph{static functions} that generate static representations of expressions and types.
 %Library providers write these static functions using the Verse \emph{static language} (SL).
-This design has conceptual roots in earlier work on \emph{active libraries}, which similarly envisioned using compile-time computation to give library providers more control over aspects of the language and compilation process \cite{active-libraries-thesis}.
+The design we are proposing also has conceptual roots in earlier work on \emph{active libraries}, which similarly envisioned using compile-time computation to give library providers more control over aspects of the language and compilation process \cite{active-libraries-thesis}.
 %Maintaining a separation between the static (or ``compile-time'') phase and the dynamic (or ``run-time'') phase is an important facet of Verse's design.
@@
 
+As vehicles for this work, we plan to formally specify small typed lambda calculi that capture each of the novel primitives that we introduce ``minimally''.
+We will also describe (but not formally specify) a new ``full-scale'' functional language called Verse.\footnote{We distinguish Verse from Wyvern, which is the language referred to in prior publications about some of the work that we will describe, because Wyvern is a group effort evolving independently in some important ways.} The reason we will not follow Standard ML \cite{mthm97-for-dart} in giving a complete formal specification of Verse is both to emphasize that the primitives we introduce can be considered for inclusion in a variety of language designs, and to avoid distracting the reader with specifications for ``orthogonal'' primitives that are already well-understood in the literature.
+%We anticipate that future full-scale language specifications will be able to combine the ideas  in the proposed work without trouble.
+%The purpose of the work being proposed is to serve as a reference for those interested in the new constructs we introduce, not to serve as a language specification.
+We will give a brief overview covering how these languages are organized in Chap. \ref{sec:verse}.
+
 The main challenge in the design of these primitives will come in ensuring that they are metatheoretically well-behaved.
-If we are not careful, many of the problems  that arise when combining language dialects, discussed earlier, could simply shift into the semantics of these primitives.\footnote{This is why languages  like Verse are often called \emph{extensible languages}, though this is somewhat of a misnomer.
+If we are not careful, many of the problems  that arise when combining language dialects, discussed earlier, could simply shift into the semantics of these primitives.\footnote{This is why languages  like Verse are often called ``extensible languages'', though this is somewhat of a misnomer.
 The defining characteristic of an extensible language is that it \emph{doesn't} need to be extended in situations where other languages would need to be extended.
@@
 
-As vehicles for this work, we plan to formally specify typed lambda calculi that capture each of the novel primitives that we introduce ``minimally''.
-We will also describe (but not formally specify) a new ``full-scale'' functional language called Verse.\footnote{We distinguish Verse from Wyvern, which is the language referred to in prior publications about some of the work being proposed here, because Wyvern is a group effort evolving independently in some important ways.} The reason we will not follow Standard ML \cite{mthm97-for-dart} in giving a complete formal specification of Verse is both to emphasize that the primitives we introduce can be considered for inclusion in a variety of language designs, and to avoid distracting the reader with specifications for ``orthogonal'' primitives that are already well-understood in the literature.
-%We anticipate that future full-scale language specifications will be able to combine the ideas  in the proposed work without trouble.
-%The purpose of the work being proposed is to serve as a reference for those interested in the new constructs we introduce, not to serve as a language specification.
-We will give a brief overview covering how these languages are organized in Sec. \ref{sec:verse}.
-
 \subsection{Thesis Statement}
@@
 A functional programming language can give library providers the ability to %meta\-pro\-gram\-matic\-ally 
-express new syntactic expansions and new type structure atop a minimal type-theoretic internal language while maintaining a hygienic type discipline and modular reasoning principles.
+express new syntactic expansions and new types and operators atop a small type-theoretic internal language while maintaining a hygienic type discipline and modular reasoning principles.
 %These  primitives are  expressive enough to subsume the need for a variety of primitives that are, or would need to be, built in to comparable contemporary languages.
@@
 
-\subsection{Motivating Example: Regular Expression Syntax}\label{sec:examples}
+\subsection{Motivating Example: Regular Expression Syntax}\label{sec:examples}\label{sec:syntax-examples}
 Let us begin by taking the perspective of a regular expression library provider.
-We assume the reader has some familiarity with regular expressions \cite{Thompson:1968:PTR:363347.363387}\todo{better ref}.
+We assume the reader has some familiarity with regular expressions \cite{Thompson:1968:PTR:363347.363387}.
 We will discuss a standard variant of regular expressions that supports marking \emph{captured groups} (with parentheses in the concrete syntax) to make certain examples more interesting.
@@
 
-\subsection{Existing Approaches}\label{sec:syntax-existing}
+\subsection{Existing Approaches}\label{sec:syntax-existing}\label{sec:syntax-existing-approaches}
 \subsubsection{Dynamic String Parsing}
@@
 
-\subsection{Motivating Example: Labeled Tuples and Regular Strings}\label{sec:metamodules-example}
+\subsection{Motivating Example: Labeled Tuples and Regular Strings}\label{sec:metamodules-example}\label{sec:metamodules-motivating-examples}
 As a simple introductory example, let us define a \emph{labeled tuple type} for conference papers:

### 2015-08-20 thesis-proposal/problem.tex
Changes in response to Bob's comments, thru Sec4

@@
 \end{enumerate} 
-Both TSMs and metamodules make extensive use of \emph{static code generation} (also called \emph{static} or \emph{compile-time metaprogramming}), meaning that the relevant rules in the static semantics of the language call for the evaluation of \emph{static functions} that generate static representations of expressions and types.
+Both TSMs and metamodules make extensive use of \emph{static code generation} (also called \emph{static} or \emph{compile-time metaprogramming}, hence the name \emph{metamodules}), meaning that the relevant rules in the static semantics of the language call for the evaluation of \emph{static functions} that generate static representations of expressions and types.
 %Library providers write these static functions using the Verse \emph{static language} (SL).
-The design we are proposing also has conceptual roots in earlier work on \emph{active libraries}, which similarly envisioned using compile-time computation to give library providers more control over aspects of the language and compilation process \cite{active-libraries-thesis}.
+The design we are proposing also has conceptual roots in earlier work on \emph{active libraries}, which similarly envisioned using compile-time computation to give library providers more control over aspects of the language and compilation process (but did not take a type theoretic approach) \cite{active-libraries-thesis}.
 %Maintaining a separation between the static (or ``compile-time'') phase and the dynamic (or ``run-time'') phase is an important facet of Verse's design.
@@
 We will avoid this somewhat confusing terminology.} Our main technical contributions will be in rigorously showing how to address these problems in a principled manner.
-In particular, syntactic conflicts will be impossible by construction and the semantics will validate code statically generated by TSMs and metamodules to maintain a strong \emph{hygienic type discipline} and, most uniquely, powerful \emph{modular reasoning principles}.
-In other words, library providers will have the ability to reason about the constructs that they have defined in isolation, and clients will be able to use them safely in any program context and in any combination, without the possibility of conflict.\footnote{This is not quite true -- simple naming conflicts can arise.
+In particular, syntactic conflicts will be impossible by construction and the semantics will validate code statically generated by TSMs and metamodules to maintain:
+\begin{itemize}
+\item a \emph{hygienic type discipline}, meaning that these constructs maintain type safety and that the validity of the code generated will not make any assumptions about the surrounding context; and
+\item \emph{modular reasoning principles}, meaning that library providers will have the ability to reason about the constructs that they have defined in isolation, and clients will be able to use them safely in any combination, without the possibility of conflict.\footnote{This is not quite true -- simple naming conflicts can arise.
 We will tacitly assume that they are being avoided extrinsically, e.g.
-by using a URI-based naming scheme as in the Java ecosystem.} We will make these notions completely precise as we continue.
+by using a URI-based naming scheme as in the Java ecosystem.}
+\end{itemize}
+We will make these notions completely precise as we continue.
 
@@
 When compiling an expression like this, the client would see an error message like \verb|error: illegal escape character|.\footnote{This is the error message that \texttt{javac} produces.
-When compiling an analagous expression using SML of New Jersey (SML/NJ), we encounter the rather bizarre error message \texttt{Error: unclosed string}.} In a small lab study, we observed that this class of error was common, and often confused even experienced programmers if they had not used regular expressions recently \cite{Omar:2012:ACC:2337223.2337324}.
+When compiling an analagous expression using SML of New Jersey (SML/NJ), we encounter a more confusing error message: \texttt{Error: unclosed string}.} In a small lab study, we observed that this class of error was common, and often confused even experienced programmers if they had not used regular expressions recently \cite{Omar:2012:ACC:2337223.2337324}.
 The standard workaround has higher syntactic cost -- we must double all backslashes:
@@
 
+\item The next problem is that dynamic string parsing mainly decreases the syntactic cost of complete patterns.
+Patterns constructed compositionally cannot easily benefit from this technique.
+For example, consider this function from strings to patterns:
+\begin{lstlisting}[numbers=none]
+fun example(name : string) => 
+  P.Seq(P.Str(name), P.Seq(pattern "SSTR: ESTR", ssn)) (* ssn as above *)
+\end{lstlisting}
+%We needed to use both dynamic string parsing and explicit applications of pattern constructors to achieve the intended semantics. 
+Had we built derived syntax for regular expression patterns into the language primitively (following Unix conventions of using forward slashes as delimiters), we could use equivalently use splicing syntax:
+\begin{lstlisting}[numbers=none]
+fun example_shorter(name : string) => /@name: %ssn/
+\end{lstlisting}
+Here, an identifier (or parenthesized expression, not shown) prefixed with an \lstinline{@} is a spliced string, and one prefixed with a \lstinline{%} is a spliced pattern.
+
+It is difficult to capture idioms like this using dynamic string parsing, because strings cannot contain sub-expressions directly.
+
+%(we will see an example of syntax that does capture such idioms below).
+
+\item For functions like \lstinline{example} where we are constructing patterns on the basis of data of type \lstinline{string}, using strings coincidentally to introduce patterns tempts programmers to use string concatenation in subtly incorrect ways.
+For example, consider the following seemingly more readable definition of \lstinline{example}:
+\begin{lstlisting}[numbers=none,escapechar=~]
+fun example_bad(name : string) => 
+  pattern (name ^ {rx|SSTR: \d\d\d-\d\d-\d\d\d\dESTR|rx})
+\end{lstlisting}
+%The (unstated) intent here was to treat \lstinline{name} as a sub-pattern matching only itself, but this is not the observed behavior when \lstinline{name} contains special characters that have other meanings in patterns.
+
+Both \lstinline{example} and \lstinline{example_bad} have the same type and behave identically at many inputs, particularly those that would be expected during typical executions of the program (i.e.
+alphabetic names).
+It is only when the input \lstinline{name} contains special characters that have meaning in the concrete syntax of patterns that a problem arises.
+
+In applications that query sensitive data, mistakes like this lead to \emph{injection attacks}, which are among the most common and catastrophic security threats on the web today \cite{owasp2013}.
+These are, of course, a consequence of the programmer making a mistake in an effort to decrease syntactic cost, but proving that mistakes like this have not been made involves reasoning about complex run-time data flows, so it is once again notoriously difficult to automate.
+If our language supported derived syntax for patterns, this kind of mistake would be substantially less common (because \lstinline{example_shorter} has lower syntactic cost than \lstinline{example_bad}).
+ %Ultimately, of course, mistakes like this are the fault of a programmer using a flawed heuristic, and they could be avoided with discipline.
+The problem is once again that it is difficult to detect violations of this discipline automatically.
+
+ %Ideally, our library would be able to make it more difficult to inadvertently introduce subtle security bugs like this.
 \item The next problem is that pattern parsing does not occur until the pattern is evaluated.
@@
 as \lstinline{"SSTR(GESTR" ^ "SSTRCESTR"} where \lstinline{^} is the string concatenation function applied in infix style, it would also need to be able to establish that this expression is equivalent to the string \lstinline{"SSTR(GCESTR"}.
-For patterns that are dynamically constructed based on input to a function (see example below), the problem is not solved by ``simply''  evaluating the expression statically (or, more generally, in some earlier ``stage'' of evaluation) \cite{Jones:Gomard:Sestoft:93:PartialEvaluation}.
+For patterns that are dynamically constructed based on input to a function, the problem is not solved by   evaluating the expression statically (or, more generally, in some earlier ``stage'' of evaluation) \cite{Jones:Gomard:Sestoft:93:PartialEvaluation}.
 
 Of course, asking the client to provide a proof of well-formedness would completely defeat the purpose -- lowering syntactic cost.
+
+In a language that primitively supported derived pattern syntax, the parsing would occur at compile-time and so malformed patterns would produce a compile-time error.
 
@@
 \end{lstlisting}
-To avoid incurring the parsing cost for each element of \lstinline{exmpl_list}, the programmer or compiler must move the parsing step out of the closure (for example, by eta-reduction in this simple example) or ensure that an appropriately tuned memoization (i.e.
-caching) strategy is being used.\footnote{Anecdotally, in scientific code using major contemporary compilers, the former optimization is not automatic.
-Compilation caches are more often provided by the regular expression library, but like any cache, it is difficult to reason compositionally about performance in their presence.} %If the programmer does it, it can sometimes make the program more difficult to read.
+To avoid incurring the parsing cost for each element of \lstinline{exmpl_list}, the programmer or compiler must move the parsing step out of the closure (for example, by eta-reduction in this simple example).\footnote{Anecdotally, in major contemporary compilers, this optimization is not automatic.} If the programmer must do this, it can (in more complex examples) increase syntactic cost and cognitive cost by moving the pattern itself far away from its use site.
+Alternatively, an appropriately tuned memoization (i.e.
+caching) strategy could be used to amortize some of this cost, but it is difficult to reason compositionally about performance using such a strategy.
+%If the programmer does it, it can sometimes make the program more difficult to read.
 
@@
 
-\item The next problem is that dynamic string parsing mainly decreases the syntactic cost of complete patterns.
-Patterns constructed compositionally cannot easily benefit from this technique.
-For example, consider this function from strings to patterns:
-\begin{lstlisting}[numbers=none]
-fun example(name : string) => 
-  P.Seq(P.Str(name), P.Seq(pattern "SSTR: ESTR", ssn)) (* ssn as above *)
-\end{lstlisting}
-%We needed to use both dynamic string parsing and explicit applications of pattern constructors to achieve the intended semantics. 
-It is difficult to capture idioms like this more concisely using dynamic string parsing (we will see an example of syntax that does capture such idioms below).
-
-\item Finally, for functions like \lstinline{example} where we are constructing patterns on the basis of data of type \lstinline{string}, using strings coincidentally to introduce patterns tempts programmers to use string concatenation in subtly incorrect ways.
-For example, consider the following seemingly more readable definition of \lstinline{example}:
-\begin{lstlisting}[numbers=none,escapechar=|]
-fun example_bad(name : string) => 
-  pattern (name ^ "SSTR: \\d\\d\\d-\\d\\d-\\d\\d\\d\\dESTR")
-\end{lstlisting}
-%The (unstated) intent here was to treat \lstinline{name} as a sub-pattern matching only itself, but this is not the observed behavior when \lstinline{name} contains special characters that have other meanings in patterns.
-
-Both \lstinline{example} and \lstinline{example_bad} have the same type and behave identically at many inputs, particularly those that would be expected during typical executions of the program (i.e.
-alphabetic names).
-It is only when the input \lstinline{name} contains special characters that have meaning in the concrete syntax of patterns that a problem arises.
-In applications that query sensitive data, mistakes like this lead to \emph{injection attacks}, which are among the most common and catastrophic security threats on the web today \cite{owasp2013}.
-Proving that mistakes like this have not been made involves reasoning about complex run-time data flows, so it is once again notoriously difficult to automate.
- %Ultimately, of course, mistakes like this are the fault of a programmer using a flawed heuristic, and they could be avoided with discipline.
-The problem is once again that it is difficult to detect violations of this discipline automatically.
-
- %Ideally, our library would be able to make it more difficult to inadvertently introduce subtle security bugs like this.
+In a language that primitively supported derived pattern syntax, the expansion is computed at compile-time and so there is no dynamic cost.
 \end{enumerate}
@@
 
-Syntax extension systems that are based on context-free grammars like  Sugar* \cite{erdweg2013framework}, Camlp4 \cite{OCaml-manual} and several others would allow us to directly introduce pattern syntax into our core language's grammar, perhaps following Unix conventions like this:
+Other syntax extension systems are more flexible.
+For example, many are based on context-free grammars, e.g.
+Sugar* \cite{erdweg2013framework}, Camlp4 \cite{ocaml-manual} and several others.
+Others give library providers direct programmatic access to the parse stream, like Common Lisp's \emph{reader macros} \cite{steele1990common} (which are distinct from its term-rewriting macros, described below) and Racket's preprocessor \cite{Flatt:2012:CLR:2063176.2063195}.
+All of these would allow us to add pattern syntax into our core language's grammar, perhaps following Unix conventions like this:
 \begin{lstlisting}[numbers=none]
@@
 
-For patterns constructed compositionally, we can define \emph{splicing syntax} that allows us to write string and pattern expressions inline (distinguished by prefixes \lstinline{@} and \lstinline{%}, respectively).
-For example, we can write \lstinline{example} equivalently, but at lower syntactic cost, like this:
-\begin{lstlisting}[numbers=none,escapechar=|]
-fun example_concise(name : string) => 
-  /@name: %ssn/
-\end{lstlisting}
+As discussed above, we can define \emph{splicing syntax} that allows us to write string and pattern expressions inline (distinguished by prefixes \lstinline{@} and \lstinline{%}, respectively). 
+For example, we can write \lstinline{example} equivalently, but at lower syntactic cost (cf. \lstinline{example_shorter}).
 %The body of this function elaborates to the body of \lstinline{example_fixed} as shown above. 
-Had we mistakenly written \lstinline{%name}, we would encounter only a static type error, rather than the  silent injection  vulnerability discussed above.
+%Had we mistakenly written \lstinline{%name}, we would encounter only a static type error, rather than the  silent injection  vulnerability discussed above.
 
@@
 If another library provider used similar syntax for a different implementation or variant of regular expressions, or for some other unrelated construct, then a client could not simultaneously use both libraries at the same time.
-So properly considered, every combination of extensions introduced using these mechanisms creates a \emph{de facto} derived syntactic dialect of our language.
-The benefit of these systems over monolithic systems like Racket's syntactic preprocessor \cite{Flatt:2012:CLR:2063176.2063195} is only that they lower the implementation cost of constructing syntactic dialects.
+So properly considered, every combination of extensions introduced using these mechanisms creates a \emph{de facto} syntactic dialect of our language.
+The benefit of these systems is only that they lower the implementation cost of constructing syntactic dialects.
 % Resolving such parsing amibiguities is left to each client of the library.
@@
 
-Putting this aside, we must also consider the question of how newly introduced forms elaborate to forms in our base grammar.
-In our example, this is tricky because we have defined a modular encoding of patterns -- which particular module should the elaboration use?
+Putting this aside, we must also consider the question of how newly introduced forms expand to forms in our base grammar.
+In our example, this is tricky because we have defined a modular encoding of patterns -- which particular module should the expansion use?
 Clearly, simply assuming that some module identified as \lstinline{P} matching \lstinline{PATTERN} is in scope is a brittle solution.
@@
 Such a \emph{hygiene discipline} is well-understood only when performing term-to-term rewriting (discussed below) or in simple language-integrated rewrite systems like those found in Coq.
-For mechanisms that operate strictly at the level of context-free grammars, this issue is not easy to address.
-
-We can address the problem of choosing a module to use in the elaboration by requiring that the client explicitly identify it as an ``argument'' to the form:
+For mechanisms that operate strictly at the level of context-free grammars or the parse stream, this issue has not been adequately addressed.
+The library provider must simply be careful not to make assumptions about variable names and instead require that the client explicitly identify it as an ``argument'' to the form:
 \begin{lstlisting}[numbers=none]
@@
 
-Another problem with the approach of direct syntax extension is that there is no typing discipline -- given an unfamiliar piece of syntax, there is no straightforward method for determining what type it will have, causing difficulties for both humans (related to code comprehension) and tools.
+Another problem with the approach of direct syntax extension is that, given an unfamiliar piece of syntax, there is no straightforward method for determining what type it will have, causing difficulties for both humans (related to code comprehension) and tools.
 
@@
 
-Here, \lstinline{pattern} is a macro parameterized by a module \lstinline{P : PATTERN}.
-It statically parses the provided string literal (which must still be written using an OCaml-style literal here) to generate an expansion.
-The macro specifies that the expansion it statically generates will be of type \lstinline{P.t}.
+The definition of \lstinline{pattern} might look like this:
+\begin{lstlisting}[numbers=none]
+macro pattern[Q : PATTERN](e) at Q.t {
+  static fun f(e :: Exp) :: Exp => case(e) {
+      StrLit(s) => (* pattern parser here *)
+    | BinOp(Caret, e1, e2) => `Q.Seq(Q.Str(%e1), %(f e2))`
+    | BinOp(Plus, e1, e2) => `Q.Seq(%(f e1), %(f e2))`
+    | _ => raise Error
+  }
+}
+\end{lstlisting}
+
+Here, \lstinline{pattern} is a macro parameterized by a module satisfying the signature \lstinline{PATTERN} (we identify it as \lstinline{Q} to emphasize that there is nothing special about the identifier \lstinline{P} that we have been using) and taking a single argument, identified as \lstinline{e}.
+The macro specifies that the expansion it statically generates will be of type \lstinline{Q.t}.
+It is defined by a static function that examines the syntax tree of the provided argument (syntax trees are of a sort \lstinline{Exp} defined in the standard library; cf.
+SML/NJ's visible compiler \cite{SML/VisibleCompiler}).
+If it is a string literal, as in the example above, it statically parses the literal body to generate an expansion (the details of the parser would be entirely standard).
 By parsing the string statically, we avoid the problems of dynamic string parsing for statically-known patterns.
@@
 
+The logic implementing this can be seen above.
+We assume that there is derived syntax available for the sort \lstinline{Exp} where the \lstinline{%} prefix indicates a spliced expression, i.e.
+\emph{quasiquotation} as in Lisp \cite{Bawd99a} and Scala \cite{shabalin2013quasiquotes}.
+
 Having to creatively repurpose existing syntax in this way limits the impact a library provider can have on syntactic cost (particularly when it would be desirable to adopt conventions that are not consistent with the conventions adopted by the language).
 It also can create confusion for readers expecting parenthesized expressions to behave in a consistent manner.
-However,  this approach is still preferably to direct syntax extension because it avoids many of the problems discussed above: there cannot be syntactic conflicts (because the syntax is not extended at all), we can define macro abbreviations because macros are integrated into the language, there is a hygiene discipline that guarantees that the expansion will not capture variables inadvertently, and using a typed macro system, there is a clear type discipline (i.e.
-programmers need not examine the expansion to know what type a macro application must have).
+However,  this approach is still preferable to direct syntax extension because it avoids many of the problems discussed above: there cannot be syntactic conflicts (because the syntax is not extended at all), we can define macro abbreviations because macros are integrated into the language, there is a hygiene discipline that guarantees that the expansion will not capture variables inadvertently, and using a typed macro system, programmers need not examine the expansion to know what type a macro application must have.
 
@@
 TSLs are TSMs associated directly with a type when it is generated.
-Verse leverages local type inference to implicitly control TSL dispatch.
+Verse uses local type inference to implicitly control TSL dispatch.
 
@@
 \begin{lstlisting}[numbers=none]
-syntax pattern(P : PATTERN) at P.t {
+syntax pattern[Q : PATTERN] at Q.t {
   static fn (ps :: ParseStream) :: Exp => (* pattern parser here *)
@@
 \end{lstlisting}
-This TSM definition first identifies the TSM as \lstinline{pattern}, then specifies a module parameter, \lstinline{P}, which must match the signature \lstinline{PATTERN}.
-Note that identifying the module parameter as \lstinline{P} here is an entirely local choice, i.e.
-the TSM can be applied to \emph{any} module parameter matching  \lstinline{PATTERN}.
-This parameter is used in the type annotation \lstinline{at P.t}, which specifies that all expansions that arise from the application of this TSM to a module \lstinline{P} and a delimited form must necessarily be of type \lstinline{P.t}.
+This TSM definition first identifies the TSM as \lstinline{pattern}, then specifies a module parameter, \lstinline{Q}, which must match the signature \lstinline{PATTERN} (again, we use \lstinline{Q} to emphasize that the library provider does not make any assumptions about identifiers in scope; the client must pass in a particular module, e.g.
+\lstinline{P} above).
+This parameter is used in the type annotation \lstinline{at Q.t}, which specifies that all expansions that arise from the application of this TSM to a module and a delimited form must necessarily be of type \lstinline{Q.t}, where \lstinline{Q} is that module.
 These expansions arise by the static action of the parse function defined next, within braces.
@@
 Both of these are defined in the Verse \emph{prelude}, which is a set of definitions available ambiently.
-The sort \lstinline{ParseStream} will give the function access to the {body} of the delimited form (in blue above) and the sort \lstinline{Exp}  encodes the abstract syntax of external expressions.
+The sort \lstinline{ParseStream} will give the function access to the {body} of the delimited form (in blue above) and the sort \lstinline{Exp}, mentioned also in the discussion of macros above,  encodes the abstract syntax of external expressions (there are also encodings of other syntactic classes, e.g.
+types).
+Note that the expansion logic cannot itself depend on the parameters (only the expansion itself can refer to the parameters).
 %So the parse function parses the body of the delimited form to generate an encoding of the elaboration.
@@
 \begin{lstlisting}[numbers=none]
-P.Seq(P.Str(<spliced>(name)), P.Seq(P.Str ": ", <spliced>(ssn)))
+Q.Seq(Q.Str(<spliced>(name)), Q.Seq(Q.Str ": ", <spliced>(ssn)))
 \end{lstlisting}
@@
 \begin{lstlisting}[numbers=none]
-P.Seq(P.Str(name), P.Seq(P.Str ": ", ssn))
+Q.Seq(Q.Str(name), Q.Seq(Q.Str ": ", ssn))
+\end{lstlisting}
+
+After checking that the expansion is valid, i.e.
+that it is hygienic and that the expansion is of the type specified by the TSM, the semantics removes the splicing markers just mentioned and substitutes  the actual module parameter \lstinline{P} for \lstinline{Q}, producing the final expansion, as expected:
+\begin{lstlisting}[numbers=none]
+P.Seq(P.Str(name), P.Seq(P.Str ": ", ssn)
 \end{lstlisting}
@@
 
-\paragraph{Formalization} To briefly introduce the formal specification of TSMs, let us consider the rule in the EL's semantics for handling TSM application to a delimited form.
+\paragraph{Formalization} To introduce the formal specification of TSMs, let us consider the rule in the EL's semantics for handling TSM application to a delimited form.
 To simplify matters, we will first consider only unparameterized TSMs (i.e.
@@
 Looking at the conclusion of the rule, we see that TSM application takes the abstract form $\mathbf{aptsm}(\psi; b)$.
-The metavariable $\psi$ ranges over \emph{TSM expressions} and $b$ ranges over {bodies} of delimited forms.
+The metavariable $\psi$ ranges over \emph{TSM expressions} and $b$ ranges over {bodies} of delimited forms (i.e.
+the sequences of characters between the delimiters in the examples above).
 TSM expressions consist only of TSM definitions here (cf.
-the definition of \lstinline{pattern} above), though when we formalize parameterized TSMs, we will distinguish TSM definitions from TSM expressions.%Note that for simplicity in the formal system, we have tacitly assumed that TSM names are drawn from a set that is disjoint from the set of variable names, so that function application and TSM application desugar to the distinct abstract forms.
+the definition of \lstinline{pattern} above), though when we formalize parameterized TSMs, we will distinguish TSM definitions from TSM expressions, which involve parameter application.%Note that for simplicity in the formal system, we have tacitly assumed that TSM names are drawn from a set that is disjoint from the set of variable names, so that function application and TSM application desugar to the distinct abstract forms.
 %In practice, we would use a common set of names and include a disambiguation step, but making the assumption of disjoint names allows us to avoid specifying this.
@@
 \subsubsection{Type-Specific Languages (TSLs)}\label{sec:tsls}
-With TSMs, we achieved the conventional syntax for regular expression patterns within delimiters, but we still explicitly had to apply the TSM and provide the module parameter each time.
-To further lower the syntactic cost of using TSMs, Verse also  supports \emph{type-specific languages} (TSLs).
-This allows library providers to associate a TSM directly with a generated type (or parameterized type constructor, more generally).
-For example, a module \lstinline{P} can associate \lstinline{pattern} with \lstinline{P.t} as follows:
-\begin{lstlisting}[numbers=none]
-module P :> PATTERN = mod {
+With TSMs, we achieved the conventional syntax for regular expression patterns within delimiters, but we still explicitly had to apply the TSM and provide the module parameter at each application site.
+To further lower the syntactic cost of using TSMs, so that it compares to the syntactic cost of derived syntax built in primitively, Verse also allows library providers to associate a TSM directly with an abstract or declared type (or type constructor, more generally).
+When the semantics encounters a delimited form not prefixed by a TSM name, the TSM associated with the type it is being analyzed against (i.e.
+that type's \emph{type-specific language}) is implicitly applied
+
+For example, a module \lstinline{P} can associate \lstinline{pattern} with \lstinline{P.t} by qualifying the definition of the opaquely ascribed module it is defined by as follows:
+\begin{lstlisting}[numbers=none]
+module P = mod {
   type t = (* ... *)
   (* ... *)
-} with syntax pattern
-\end{lstlisting}
-
-When the semantics encounters a delimited form not prefixed by a TSM name, the TSM associated with the type it is being analyzed against is implicitly applied.
-For example, the following function is equivalent to \lstinline{example_tsm} (note the return type annotation):
+} :> PATTERN with syntax pattern
+\end{lstlisting}
+This definition associates the TSM \lstinline{pattern P} with the abstract type \lstinline{P.t}.
+As such, the following function is equivalent to \lstinline{example_tsm} (note the return type annotation, which is now necessary):
 \begin{lstlisting}[numbers=none]
@@
 From the client's perspective, it is essentially as if the language had built in derived syntax for lists and regular expression patterns directly.%However, we did not need to build in this syntax primitively.%The only constraint is that this syntax must be used in an analytic position, which we argue is actually better for code compren when encountering unfamiliar syntax.
+
+\paragraph{Abstract Types}
+Let us examine how this mechanism works for abstract types in more detail.
+When sealing a module expression with a signature, the programmer can specify a previously defined TSMs for each abstract type that arises from the ascription.
+Each must be parameterized by a module satisfying the specified signature.
+
+For parameterized modules, the situation is similar. For example, consider a trivially parameterized module that creates modules satisfying \lstinline{PATTERN}:
+\begin{lstlisting}[numbers=none]
+module F() => mod {
+  type t = (* ... *)
+  (* ... *)
+} :> PATTERN with syntax pattern 
+\end{lstlisting}
+Each application of \lstinline{F} generates a distinct abstract type. The semantics associates the appropriately parameterized TSM with each of these:
+\begin{lstlisting}[numbers=none]
+module F1 = F() (* F1.t has TSL pattern(F1) *)
+module F2 = F() (* F2.t has TSL pattern(F2) *)
+\end{lstlisting}
+
+For a parameterized module that takes non-trivial parameters, we need only define the TSM to also take the same parameters and apply them explicitly within the definition of the parameterized module.
+For example, let us define a parameterized module \lstinline{G : A -> B}:
+\begin{lstlisting}[numbers=none,mathescape=|]
+signature A = sig { type t; val x : t }
+signature B = sig { type u; val y ; u }
+syntax $G(M : A)(G : B) at G.u { (* ... *) }
+module G(M : A) => mod { 
+  type u = M.t; val y = M.x } :> B with syntax $G(M)
+\end{lstlisting}
+Note that within the definition of \lstinline{G}, type \lstinline{u} does not have a TSL available to it.
+
+TSL lookup respects type equality, so any aliases to a type with a TSL will also have that TSL.
+We can see this in the following example, where the type \lstinline{u} has a different TSL associated with it inside and outside the definition of the module \lstinline{N}:
+\begin{lstlisting}[numbers=none,mathescape=|]
+module M : A = mod { type t = int; val x = 0 }
+module G1 = G(M) (* G1.t has TSL $G(M), per above *)
+module N = mod { 
+  type u = G1.t (* u = G1.t in this scope, so u also has TSL $G(M) *)
+  val y = /asdf/ (* we can use it to create a value of that type *) 
+} :> B (* did not specify a TSL for N.u at the point where it is sealed, 
+            so N.u has no TSL in this scope *)
+\end{lstlisting}
 
@@
 
-Another concern is that if we define the SL such that evaluating the parse function may not terminate, this then causes typechecking to become undecidable.
-In practice, this too should  be a minor concern.
-For example, we could specify a limit to the running time of the parse function.
-The compiler would enforce this limit to ensure that compilation does not ``hang'' due to an incorrect parser.
-Related concerns would arise if the static language supported external (e.g.
-I/O) effects.
-Existing techniques for controlling these effects would handle this problem, but this is not something we plan to detail here because our SL does not specify these effects.
+To provide the strongest metatheoretic guarantees, we must have that the SL is a total functional language.
+If we instead defined the SL such that evaluating the parse function may not terminate, this would then cause typechecking to become undecidable.\footnote{A ``hack'' that could be used in practice is to specify a limit to the running time of the parse function.
+The compiler would enforce this limit to ensure that compilation does not ``hang'' due to an incorrect parser.} Related concerns would arise if the static language supported external (e.g.
+I/O) effects -- one would generally not want the mere act of parsing and typechecking a program to have an effect on a file system.
+Mutation effects could be introduced, however, because the static language is kept distinct from the external language.
+No static values can be ``promoted'' into the external language.
 
@@
 control flow idioms.
-In fact, this is intentional, as it makes maintaining a type discipline in the presence of TSMs substantially easier -- neither humans nor tools need to examine the expansion to determine what type it must necessarily have.
+In fact, this is intentional -- neither humans nor tools need to examine the expansion to determine what type it must necessarily have.
 We will, however, discuss other points in the design space in the dissertation.
@@
 In fact, we conjecture that it is completely straightforward to extend this work to allow a TSM to define both an expression parser and a pattern parser.
-We plan to informally discuss this as the most immediate avenue for future work, but will not consider pattern matching in formal detail in the dissertation.
+Similarly, one could imagine defining TSMs at the level of modules, and at the level of types.
+We plan to discuss these in the dissertation.
+Formally specifying these will be left to future work, as it would introduce a rather large amount of technical overhead for minimal conceptual gain.
+%, the essential ideas are mostly simply conveyed by sticking to formal specifications only at the level of expressions.
 
@@
 The rule (syn-aptsm) above represents the first and most substantial step of this plan.
-I anticipate that this will take about \textbf{1.5 weeks} of further work.
+This is complete on paper, so I only need to typeset it.
 
 Wyvern does not specify an ML-style module system (type declarations are the only form of type generation) and at the time, it did not support parameterized types, so to actually support the examples we have given in this section, we then need to update the formalization to handle parameterized TSMs.
-I plan to flesh this out in the course of writing the dissertation, and I anticipate it requiring 5.5 weeks of my time after the proposal is complete.
+I plan to flesh this out in the course of writing the dissertation, and I anticipate it requiring a further 1 month of my time after the proposal is complete.
 I propose the following concrete milestones:
@@
 \begin{enumerate}
-\item equip the EL with an appropriately structured module context \textbf{(2 weeks)}
-\item formally specify the syntax of parameterized TSM expressions \textbf{(0.5 weeks)}
-\item formally specify an updated version of the TSM expression validation judgement that appeared as the first premise of (syn-aptsm) above \textbf{(1 week)}
-\item update the metatheory established in the work cited above to handle this \textbf{(2 weeks)}
+\item equip the EL with an appropriately structured module context \textbf{(complete on paper, needs to be typeset)}
+\item formally specify the syntax of parameterized TSM expressions \textbf{(complete on paper, needs to be typeset)}
+\item formally specify an updated version of the TSM expression validation judgement that appeared as the first premise of (syn-aptsm) above \textbf{(complete on paper, needs to be typeset)}
+\item update the metatheory established in the work cited above to handle this
 \end{enumerate}
@@
 I will also briefly summarize the corpus analysis we conducted in \cite{TSLs} where we found a large number of  situations in large publicly available Java libraries where dynamic string parsing was apparently being used to control syntactic cost.
-I anticipate that writing these examples will take an additional \textbf{1 week} of work, bringing the total for this portion of the dissertation up to 8 weeks.
+I anticipate that writing these examples will take an additional 1 week of work, bringing the total for this portion of the dissertation up to about 5 weeks of writing (plus time for review by the committee).
 %Adding in the time I will need to propose and some buffer time for unanticipated issues, my goal is to have this portion of the dissertation complete by the end of Summer 2015.

### 2015-09-12 thesis-proposal/problem.tex
Updates per Bob's comments

@@
 One way to express this abstract syntax is by defining a recursive sum type \cite{pfpl}.
-Verse supports these as case types, which are comparable to datatypes in ML (we plan to show how the type structure  of case types can in fact be expressed in libraries later):
-
-\begin{lstlisting}[numbers=none]
-casetype Pattern {
+Verse supports these as datatypes, which are comparable to datatypes in ML (we plan to show how the type structure  of datatypes can in fact be expressed in libraries later):
+
+\begin{lstlisting}[numbers=none]
+datatype Pattern {
     Empty | Str of string | Seq of Pattern * Pattern 
@@
 TSMs defined at a single type, e.g.
-the case type \lstinline{Pattern}, rather than a parameterized family of types):
+the  datatype \lstinline{Pattern}, rather than a parameterized family of types):
 \begin{mathpar}
@@
 
-As another example, we can use TSLs to express derived list syntax. For example, if we use a case type, we can define the TSL directly upon declaring the list type:
-\begin{lstlisting}[numbers=none]
-casetype list('a) { Nil | Cons of 'a * list('a) } with syntax {
+As another example, we can use TSLs to express derived list syntax. For example, if we use a datatype, we can define the TSL directly upon declaring the list type:
+\begin{lstlisting}[numbers=none]
+datatype list('a) { Nil | Cons of 'a * list('a) } with syntax {
   static fn (body :: ParseStream) => 
@@
 signature A = sig { type t; val x : t }
-signature B = sig { type u; val y ; u }
+signature B = sig { type u; val y : u }
 syntax $G(M : A)(G : B) at G.u { (* ... *) }
@@
 To provide the strongest metatheoretic guarantees, we must have that the SL is a total functional language.
-If we instead defined the SL such that evaluating the parse function may not terminate, this would then cause typechecking to become undecidable.\footnote{A ``hack'' that could be used in practice is to specify a limit to the running time of the parse function.
-The compiler would enforce this limit to ensure that compilation does not ``hang'' due to an incorrect parser.} Related concerns would arise if the static language supported external (e.g.
+If we instead defined the SL such that evaluating the parse function may not terminate, this would then cause typechecking to become undecidable.\footnote{This source of undecidability might not be particularly important in practice, for the same reason that non-termination is tolerated in most major languages.} Related concerns would arise if the static language supported external (e.g.
 I/O) effects -- one would generally not want the mere act of parsing and typechecking a program to have an effect on a file system.
@@
 
-Finally, we have discussed only expression syntax here, but have not considered pattern matching.
+We have discussed only expression syntax here, but have not considered pattern matching.
 In fact, we conjecture that it is completely straightforward to extend this work to allow a TSM to define both an expression parser and a pattern parser.
@@
 In the previous section, we assumed that the EL had an otherwise standard, ML-like type structure and considered only TSMs and TSLs as departures from ML.
-In fact, constructs like case types (shown briefly in use in the previous section), tuples, records and objects are not built primitively into Verse.
+In fact, constructs like datatypes (shown briefly in use in the previous section), tuples, records and objects are not built primitively into Verse.
 Instead, they can be expressed using \emph{metamodules} (which are distinct from, though closely related to, {modules}, as we will discuss below).
@@
 \subsection{Motivating Example: Labeled Tuples and Regular Strings}\label{sec:metamodules-example}\label{sec:metamodules-motivating-examples}
-As a simple introductory example, let us define a \emph{labeled tuple type} for conference papers:
+As a simple introductory example showing the sorts of types and operators we might like to have available to us as clients of a language, let us define a simple \emph{labeled tuple type} for conference papers:
 \begin{lstlisting}[numbers=none]
@@
 it cannot be given function type, because it is able to operate on values of \emph{any} labeled tuple type that has a component labeled \lstinline{conf}, not just \lstinline{Paper}.
-That type's component specification determines the type of the operation as a whole, here \lstinline{rstring /SURL([A-Z]+) (\d\d\d\d)EURL/}.
+The argument type's component specification determines the type of the operation as a whole, here \lstinline{rstring /SURL([A-Z]+) (\d\d\d\d)EURL/}.
 
@@
 \end{lstlisting}
-There are many other operators that we might wish to describe, for both labeled tuples and regular strings, but for our purposes it suffices to stop here.
+There are many other operators that we might wish to describe, for both labeled tuples \cite{Cardelli91operationson} and regular strings \cite{sanitation-psp14}, but for our purposes it suffices to stop here.
 
@@
   val intro : title_t -> conf_t -> t
-  (* introduction with explicit labels, in both orders *)
+  (* introduction with explicit labels, in primary order *)
   val intro_title_conf : title_t -> conf_t -> t
-  val intro_conf_title : conf_t -> title_t -> t
   (* projection *)
@@
 Here, we've taken all possible valid introductory and projection operators and called for their expression as functions.
-Expressing labels explicitly is an important facet of having labeled tuple types in a language, so we must put them into the function identifiers.
+Expressing labels explicitly is an important facet of having labeled tuple types in a language, so we put them into the function identifiers.
+This is, however, suboptimal because in client code, the argument values will not appear adjacent to the corresponding labels.
+
 Even with our minimal EL, we could  implement the semantics of the corresponding operators against this signature using a Church-style encoding (the details are standard and omitted here).
@@
    fun intro_title_conf title conf => (title, conf)
-   fun intro_conf_title conf title => (title, conf)
    fun prj_title (title, conf) => title 
@@
 }{
-	\vdash_\Phi \mathbf{ty}(m.c;\sigma_\text{param})~\mathtt{type}\leadsto[{\delta}]\tau_\text{abstrans}
+	\vdash_\Phi \mathbf{ty}(m\cdot c;\sigma_\text{param})~\mathtt{type}\leadsto[{\delta}]\tau_\text{abstrans}
 }
@@
 We plan to detail these in the thesis.
-On the other hand, constructs like ML-style exceptions, OCaml-style open datatypes or Java-style class-based object systems would not be possible because they require statically tracking which constructors/classes are in scope.
+On the other hand, constructs like ML-style exception values, OCaml-style open datatypes or Java-style class-based object systems would not be possible because they require statically tracking which constructors/classes are in scope.
 We plan to discuss adding support for new user-defined static contexts as an avenue for future work in the dissertation.
@@
 
-To further validate our work, we plan to detail a number of the examples mentioned above, including regular strings, case types, and a simple prototypic object system.
+To further validate our work, we plan to detail a number of the examples mentioned above, including regular strings, datatypes, and a simple prototypic object system.
 We anticipate that the writing involved to do this, along with other writing tasks, will take another \textbf{2 weeks}.

### 2015-09-22 thesis-proposal/problem.tex
Addressed rwh's comments, round 2

@@
 A formal account of such a system has been developed by Griffin \cite{5134}.
-Unfortunately, these systems are not expressive enough to allow us to express pattern syntax in the conventional manner.
-For example, sequences of characters  can only be parsed as identifiers using these systems, rather than as characters in a regular expression pattern.
+Unfortunately, a single equation is not enough to allow us to express pattern syntax following the usual conventions.
+For example, there is no way to handle sequences of characters (they can only be treated as variables) or escape characters (there is no way to define disjunctive syntax) using a system like Coq's.
 
@@
 Sugar* \cite{erdweg2013framework}, Camlp4 \cite{ocaml-manual} and several others.
-Others give library providers direct programmatic access to the parse stream, like Common Lisp's \emph{reader macros} \cite{steele1990common} (which are distinct from its term-rewriting macros, described below) and Racket's preprocessor \cite{Flatt:2012:CLR:2063176.2063195}.
+Other systems give library providers direct programmatic access to the parse stream, like Common Lisp's \emph{reader macros} \cite{steele1990common} (which are distinct from its term-rewriting macros, described below) and Racket's preprocessor \cite{Flatt:2012:CLR:2063176.2063195}.
 All of these would allow us to add pattern syntax into our core language's grammar, perhaps following Unix conventions like this:
@@
 
-The problems of dynamic string parsing described above are solved by this approach.
-Unfortunately, it introduces serious new problems.
-First, the systems mentioned thus far cannot guarantee that \emph{syntactic conflicts} between such extensions will not arise.
+We sidestep the problems of dynamic string parsing described above  if we directly extend the syntax of our language using any of these approaches.
+Unfortunately, direct syntax extension introduces serious new problems.
+First, the systems mentioned thus far cannot guarantee that {syntactic conflicts} between such extensions will not arise.
 As stated directly in the  Coq manual: ``mixing different symbolic notations in [the] same text may cause serious parsing ambiguity''.
@@
 
-In response to this problem, Schwerdfeger and Van Wyk developed a modular analysis that accepts only grammar extensions that specify a universally unique starting token and obey subtle constraints on  the follow sets of base language non-terminals \cite{conf/pldi/SchwerdfegerW09}.
+In response to this problem, Schwerdfeger and Van Wyk developed a modular analysis that accepts only grammar extensions that specify a unique starting token and obey subtle constraints on  the follow sets of base language non-terminals \cite{conf/pldi/SchwerdfegerW09}.
 Extensions that satisfy these criteria can be used together in any combination without the possibility of syntactic conflict.
-However, note that the most natural starting tokens like \lstinline{pattern} cannot be guaranteed to be universally unique, so we would be forced to use a more verbose token like \lstinline{edu_cmu_verse_rx_pattern}.
-There is no simple way for clients of our extension to define scoped abbreviations for starting tokens because this mechanism operates purely on syntactic forms.
+However, the most natural starting tokens like \lstinline{pattern} cannot be guaranteed to be unique.
+To address this problem, programmers must agree on a convention for defining ``globally unique identifiers'', e.g.
+the common URI convention used on the web and by the Java ecosystem.
+This forces us to use a more verbose token like \lstinline{edu_cmu_verse_rx_pattern}.
+There is no simple way for clients of our extension to define scoped abbreviations for starting tokens because this mechanism operates purely at the level of syntactic forms.
 
@@
 For mechanisms that operate strictly at the level of context-free grammars or the parse stream, this issue has not been adequately addressed.
-The library provider must simply be careful not to make assumptions about variable names and instead require that the client explicitly identify it as an ``argument'' to the form:
+The library provider must simply be careful not to make assumptions about variable names and instead require that the client explicitly identify the module they intend to use as an ``argument'' in the new form:
 \begin{lstlisting}[numbers=none]
@@
 
-Having to creatively repurpose existing syntax in this way limits the impact a library provider can have on syntactic cost (particularly when it would be desirable to adopt conventions that are not consistent with the conventions adopted by the language).
+Having to creatively repurpose existing syntax in this way limits the effect a library provider can have on syntactic cost (particularly when it would be desirable to adopt conventions that are not consistent with the conventions adopted by the language).
 It also can create confusion for readers expecting parenthesized expressions to behave in a consistent manner.
-However,  this approach is still preferable to direct syntax extension because it avoids many of the problems discussed above: there cannot be syntactic conflicts (because the syntax is not extended at all), we can define macro abbreviations because macros are integrated into the language, there is a hygiene discipline that guarantees that the expansion will not capture variables inadvertently, and using a typed macro system, programmers need not examine the expansion to know what type a macro application must have.
+However,  this approach is still preferable to direct syntax extension because it avoids many of the problems discussed above: there cannot be syntactic conflicts (because the syntax is not extended at all), we can define macro abbreviations because macros are integrated into the language, there is a hygiene discipline that guarantees that the expansion will not capture variables inadvertently, and by using a typed macro system, programmers need not examine the expansion to know what type a macro application must have.
 
@@
 \begin{lstlisting}[numbers=none]
-syntax pattern[Q : PATTERN] at Q.t {
-  static fn (ps :: ParseStream) :: Exp => (* pattern parser here *)
+syntax pattern(Q : PATTERN) at Q.t {
+  static fn (body :: Body) :: Exp => (* pattern parser here *)
 }
@@
 \lstinline{P} above).
-This parameter is used in the type annotation \lstinline{at Q.t}, which specifies that all expansions that arise from the application of this TSM to a module and a delimited form must necessarily be of type \lstinline{Q.t}, where \lstinline{Q} is that module.
-These expansions arise by the static action of the parse function defined next, within braces.
-It is written as a static function of sort \lstinline{ParseStream -> Exp}.
-Both of these are defined in the Verse \emph{prelude}, which is a set of definitions available ambiently.
-The sort \lstinline{ParseStream} will give the function access to the {body} of the delimited form (in blue above) and the sort \lstinline{Exp}, mentioned also in the discussion of macros above,  encodes the abstract syntax of external expressions (there are also encodings of other syntactic classes, e.g.
-types).
-Note that the expansion logic cannot itself depend on the parameters (only the expansion itself can refer to the parameters).
-%So the parse function parses the body of the delimited form to generate an encoding of the elaboration.
+The type annotation \lstinline{at Q.t} specifies that all expansions that arise from the application of this TSM to a module and a delimited form must necessarily be of type \lstinline{Q.t}, where \lstinline{Q} refers to that module.
+
+When a client applies the TSM to the necessary parameters and a delimited form, e.g.
+the parameter \lstinline{P} and the delimited form \lstinline{/SURLA|T|G|CEURL/} in the example above, the expansion is computed by calling the \emph{parse function} defined within braces.
+This parse function is a static function of sort \lstinline{Body -> Exp}.
+Both of these sorts are defined in the Verse \emph{prelude}, which is a set of definitions available ambiently.
+The sort \lstinline{Body} gives the static function access to the {body} of the delimited form (e.g.
+the characters in blue above) and the sort \lstinline{Exp}, mentioned also in the discussion of macros above,  encodes the abstract syntax of external expressions (there are also encodings of other syntactic classes, e.g.
+types, that can appear in expressions).
+The parse function must treat the TSM parameters parametrically, i.e.
+it does not have access to any values in the supplied module parameter.
+Only the expansion the parse function generates can refer to module parameters.
+%For example, the following definition is ill-sorted:
+%\begin{lstlisting}[numbers=none]
+%syntax pattern_bad[Q : PATTERN] at Q.t {
+%  static fn (body :: Body) :: Exp => 
+%    if Q.flag then (* ... *) else (* ... *)
+%}
+%\end{lstlisting}%So the parse function parses the body of the delimited form to generate an encoding of the elaboration.
 
 To support splicing syntax as described in Sec.
-\ref{sec:syntax-extension}, the parse function must be able to extract external subexpressions directly from the parse stream.
+\ref{sec:syntax-extension}, the parse function must be able to extract external subexpressions directly from the supplied body.
 For example, consider the client code below:
@@
 \end{lstlisting}
-The subexpressions \lstinline{name} and \lstinline{ssn} on the last line occur directly in the parse stream.
-When the parse function encounters these, it asks the \lstinline{ParseStream} to extract these as \emph{spliced expressions} for use in the expansion.
+The subexpressions \lstinline{name} and \lstinline{ssn} on the last line appear directly in the body of the delimited form, so we call them \emph{spliced subexpressions}.
+When the parse function determines that a subsequence of the body should be treated as a spliced subexpression, it can mark it as such and insert it directly into the expansion.
 For example, the expansion generated for the body of \lstinline{example_tsm} above would, if written concretely with spliced expressions marked, be:
@@
 The hygiene mechanism then ensures that only these marked portions of the generated expansion can refer to the variables at the use site, preventing inadvertent variable capture by the expansion.
-For example, the following would not be a valid expansion, because it has free variables that are not inside spliced subexpressions:
-\begin{lstlisting}[numbers=none]
-Q.Seq(Q.Str(name), Q.Seq(Q.Str ": ", ssn))
-\end{lstlisting}
+%For example, had the  would not be a valid expansion, because the  that are not inside spliced subexpressions:
+%\begin{lstlisting}[numbers=none]
+%Q.Seq(Q.Str(name), Q.Seq(Q.Str ": ", ssn))
+%\end{lstlisting}
 
@@
 \begin{lstlisting}[numbers=none]
-P.Seq(P.Str(name), P.Seq(P.Str ": ", ssn)
-\end{lstlisting}
+P.Seq(P.Str(name), P.Seq(P.Str ": ", ssn))\end{lstlisting}
 %Put another way, the  elaboration logic must be valid in any context.
@@
     \vdash \psi ~@~ \sigma_\text{ty}~\{\sigma_\text{parser}\}\\
-    \mathbf{parsestream}(b) = \sigma_\text{ps}\\
-    \sigma_\text{parser}(\sigma_\text{ps}) \Downarrow \sigma_\text{exp}\\
+    b \downarrow \sigma_\text{body}\\
+    \sigma_\text{parser}(\sigma_\text{body}) \Downarrow \sigma_\text{exp}\\
     \sigma_\text{exp} \uparrow e_\text{exp}\\\\
@@
 % Application of a parameterized TSM to its parameters is handled by this judgement (not shown).
-\item The second premise creates a parse stream, $\sigma_\text{ps}$, of sort $\mathsf{ParseStream}$, from the body of the delimited form.
+\item The second premise creates an encoding of the body of the delimited form, $\sigma_\text{body}$, of sort $\mathsf{Body}$.
 \item The third premise applies the parse function to the parse stream to generate a static representation of the expansion, $\sigma_\text{exp}$ (of sort $\mathsf{Exp}$).% Note that this is ``metaprogramming'' -- a static expression in the external semantics to generate an encoding of another external term.
@@
 
-For parameterized modules, the situation is similar. For example, consider a trivially parameterized module that creates modules satisfying \lstinline{PATTERN}:
+For parameterized modules, the situation is similar. For example, consider a trivially parameterized module that creates modules sealed against \lstinline{PATTERN}:
 \begin{lstlisting}[numbers=none]
@@
 
-TSL lookup respects type equality, so any aliases to a type with a TSL will also have that TSL.
+TSL lookup respects type equality, so any synonyms of a type with a TSL will also have that TSL.
 We can see this in the following example, where the type \lstinline{u} has a different TSL associated with it inside and outside the definition of the module \lstinline{N}:
@@
 } :> B (* did not specify a TSL for N.u at the point where it is sealed, 
-            so N.u has no TSL in this scope *)
+            so N.u has no TSL in the outer scope *)
 \end{lstlisting}
@@
 To provide the strongest metatheoretic guarantees, we must have that the SL is a total functional language.
-If we instead defined the SL such that evaluating the parse function may not terminate, this would then cause typechecking to become undecidable.\footnote{This source of undecidability might not be particularly important in practice, for the same reason that non-termination is tolerated in most major languages.} Related concerns would arise if the static language supported external (e.g.
-I/O) effects -- one would generally not want the mere act of parsing and typechecking a program to have an effect on a file system.
-Mutation effects could be introduced, however, because the static language is kept distinct from the external language.
+Allowing non-termination in the SL would cause typechecking of the EL to become undecidable because macro expansion occurs during typechecking (typechecking the IL would remain decidable, however).
+Related concerns would arise if the static language supported external (e.g.
+I/O) effects -- one would generally not want the mere act of parsing and typechecking a program to have an arbitrary effect on, for example, the programmer's file system.
+Mutation effects could, however, be introduced, because the static language is distinct from the external language.
 No static values can be ``promoted'' into the external language.
@@
 % }{
-% 	\hat{\Gamma} \vdash_\Phi \mathbf{opconap}(m; \text{op}; \sigma_\text{param}; \bar{e}) \Leftarrow \sigma_\text{ty} \leadsto [\delta][\gamma]\iota
+% 	\hat{\Gamma} \vdash_\Phi \mathbf{opconap}(m; \text{op}; \sigma_\text{param}; \dot{e}) \Leftarrow \sigma_\text{ty} \leadsto [\delta][\gamma]\iota
 % }

### 2015-09-24 thesis-proposal/problem.tex
Started removing metamodules stuff from proposal.

@@
 Functional programming language designers have long studied small typed lambda calculi to develop a principled understanding of  fundamental metatheoretic issues, like type safety, and to examine the mathematical character of language primitives of interest in isolation.
-These studies have then informed the design of ``full-scale''\footnote{Throughout this work, phrases that should be read as having an intuitive meaning, rather than a strict mathematical meaning, will be first introduced with quotation marks.} functional languages, which combine several such primitives and also typically feature various generalizations and primitive ``embellishments''  motivated by a consideration of human factors.
-For example, major languages like Standard ML (SML) \cite{mthm97-for-dart,harper1997programming}, OCaml \cite{ocaml-manual} and Haskell \cite{jones2003haskell} all primitively build in record types, generalizing the nullary and binary product types that suffice in simpler calculi, because explicitly labeled components are cognitively useful to human programmers.
+These studies have informed the design of ``full-scale''\footnote{Throughout this work, phrases that should be read as having an intuitive meaning, rather than a strict mathematical meaning, will be introduced with quotation marks.} functional languages, which combine several such primitives and also introduce various generalizations and primitive ``embellishments''  motivated by a consideration of human factors.
+For example, major functional languages like Standard ML (SML) \cite{mthm97-for-dart,harper1997programming}, OCaml \cite{ocaml-manual} and Haskell \cite{jones2003haskell} all primitively build in record types, generalizing the nullary and binary product types that suffice in simpler calculi, because explicitly labeled components are cognitively useful to human programmers.
 Similarly, these languages build in derived syntactic forms (colloquially, ``syntactic sugar'') that decrease the syntactic cost of  working with common library constructs, like lists.
@@
 The hope amongst many language designers is that a limited number of primitives like these will suffice to produce a ``general-purpose'' programming language, i.e.
-one where programmers can direct their efforts almost exclusively toward the development of useful libraries.
+one where programmers can direct their efforts almost exclusively toward the development of libraries for a wide variety of application domains.
 However, a stable language design that fully achieves this ideal has yet to emerge, as evidenced by the diverse array of ``dialects'' that continue to proliferate around all major contemporary languages.
 In fact, tools that aid in the construction of so-called  ``domain-specific'' language dialects (DSLs)\footnote{In some parts of the literature, such dialects are called ``external DSLs'', to distinguish them from  ``internal'' or ``embedded DSLs'', which are actually  library interfaces that only ``resemble'' distinct dialects \cite{fowler2010domain}.} seem only to be becoming increasingly prominent.
-
-\subsection{Why are there so many dialects?}
+%\subsection{Why are there so many dialects?}
 {This calls for an investigation}: why is it that programmers and researchers are still so often unable to satisfyingly express the constructs that they need in libraries, as modes of use of the ``general-purpose'' primitives already available in major languages today, and instead see a need for new language dialects?
 
-Perhaps the most common motivation may be that the \emph{syntactic cost} of expressing a construct of interest using contemporary general-purpose primitives is not always ideal.
+Some of these dialects extend the semantics of the language that they are based on, but the more common situation is that the existing semantic primitives suffice and it is only the \emph{syntactic cost} of expressing a construct of interest using these primitives  that is too high.
 In response, library providers construct \emph{syntactic dialects} -- dialects that introduce only new derived syntactic forms.
@@
 %Syntactic cost is often assessed qualitatively \cite{green1996usability}, though quantitative metrics can be defined.
-These are certainly not the only types of data that could similarly benefit from the availability of specialized derived forms -- we will consider another example in Sec. \ref{sec:syntax-examples}. 
+This is not an isolated example -- there are many other types of data that could similarly benefit from the availability of specialized derived forms (we will consider another example, regular expression patterns, in Sec.
+\ref{sec:syntax-examples}).
 Tools like Camlp4 \cite{ocaml-manual}, Sugar* \cite{erdweg2011sugarj,erdweg2013framework} and Racket \cite{Flatt:2012:CLR:2063176.2063195}, which we will discuss in Sec.
-\ref{sec:syntax-existing-approaches}, have lowered the engineering costs of constructing syntactic dialects in such situations, contributing to their proliferation.
-
-More advanced dialects introduce new type structure, going beyond what is possible with only new derived forms.
+\ref{sec:syntax-existing-approaches}, have lowered the engineering costs of constructing syntactic dialects in such situations, further contributing to their proliferation.
+
+%More advanced dialects introduce new type structure, going beyond what is possible with only new derived forms.
 As a simple example, the static and dynamic semantics of records cannot be expressed by context-independent expansion to a language with only nullary and binary products.
 Various languages have explored ``record-like'' primitives that go further, supporting functional update operators, width and depth coercions (sometimes implicit)%\cite{Cardelli:1984:SMI:1096.1098}
-, methods, prototypic dispatch and other such ``semantic embellishments'' that in turn cannot be expressed by context-independent expansion to a language with only standard record types (we will detail an  example in Sec.
+%, methods, prototypic dispatch and other such ``semantic embellishments'' that in turn cannot be expressed by context-independent expansion to a language with only standard record types (we will detail an  example in Sec.
 \ref{sec:metamodules-motivating-examples}).
@@
 Manticore \cite{conf/popl/FluetRRSX07} and AliceML  \cite{AliceLookingGlass} build in parallel programming primitives with a more elaborate type structure than is found in simpler accounts of parallelism.
-MLj builds in the type structure of the Java object system (motivated by a desire to interface safely and naturally with Java libraries) \cite{Benton:1999:IWW:317636.317791}.
+%MLj builds in the type structure of the Java object system (motivated by a desire to interface safely and naturally with Java libraries) \cite{Benton:1999:IWW:317636.317791}.
 Other dialects do the same for other foreign languages, e.g.
@@
 % test 
-Some have viewed the ongoing proliferation of dialects described above as harmless or even as desirable, arguing that programmers can simply choose the right language for the job \cite{journals/stp/Ward94}.
-However, this ``dialect-oriented'' approach is, in an important sense, anti-modular: a library written in one dialect cannot, in general, safely and idiomatically interface with a library written in another dialect.
-As the study of MLj demonstrated, addressing this interoperability problem requires somehow ``combining'' the  dialects into a single language.
+Some  view the ongoing proliferation of dialects described above as harmless or even as desirable, arguing that programmers can simply choose the right dialect for the job at hand \cite{journals/stp/Ward94}.
+However, this ``dialect-oriented'' approach is, in an important sense, anti-modular: a library written in one dialect cannot, in general, safely and idiomatically interface with a library written using another dialect.
+Addressing this interoperability problem requires somehow ``combining'' the  dialects into a single language \cite{Benton:1999:IWW:317636.317791}.
 However, in the most general setting where the dialects in question might be specified by judgements of arbitrary form, this is not a well-defined notion.
-Even if we restrict our interest to dialects specified using formalisms that do operationalize some notion of dialect combination, there is generally no guarantee that the combined dialect will conserve important syntactic and semantic properties that can be established about the dialects in isolation.
+Even if we restrict our interest to dialects specified using formalisms that do operationalize some notion of dialect combination, there is generally no guarantee that the combined dialect will conserve important metatheoretic properties that can be established about the dialects in isolation.
 %In other words, any putative ``combined language'' must formally be considered a  distinct system for which one must derive essentially all metatheorems of interest anew, guided only informally by those derived for the dialects individually.
@@
 This \emph{ad hoc} approach is not sustainable, for three main reasons.
-First, as suggested by the diversity of examples given above, there are simply too  many potentially useful such primitives, and many of these are only relevant in relatively narrow application domains (for derived syntax, our group has  gathered initial data speaking to this \cite{TSLs}).
+First,  there are simply too  many potentially useful such primitives, and many of these are only relevant in relatively narrow application domains (for derived syntax, our group has  gathered initial data speaking to this \cite{TSLs}).
 Second, primitives introduced earlier in a language's lifespan can end up monopolizing finite ``syntactic resources'', forcing subsequent primitives to use ever more esoteric forms.
-And third, primitives that prove to be flawed in some way cannot be removed or changed without breaking backwards compatibility.
+And third, primitives that prove afteer some time to be flawed in some way cannot be removed or replaced without breaking backwards compatibility.
 %Because there is no data about how useful a construct is in practice until it is available in a major language, decisions about which constructs to include are often informed only by intuition.
@@
 
-This suggests that language designers should be quite careful and strive to keep general-purpose languages small and free of \emph{ad hoc} primitives.
-Given that we also want to avoid dialect formation, for the reasons described above, this leaves  two possible paths forward.
-One, exemplified (arguably) by SML, is to simply eschew further syntactic conveniences and innovative type structure and settle on the existing primitives, which might be considered to sit at a ``sweet spot'' in the overall language design space.
+This suggests that language designers should strive to keep general-purpose languages small, stable and free of \emph{ad hoc} primitives.
+This leaves  two possible paths forward.
+One, exemplified (arguably) by SML, is to simply eschew further ``embellishments'' and settle on the existing primitives, which might be considered to sit at a ``sweet spot'' in the overall language design space.
 % (accepting that in some circumstances, this trades away expressive power or leads to  high syntactic cost).
-The other path forward is to search for a small number of highly general primitives that allow us degrade many of the constructs that are built primitively into languages today instead to modularly composable library constructs.
+The other path forward is to search for a small number of highly general primitives that allow us degrade many of the constructs that are built primitively into dialects today instead to modularly composable library constructs.
 Encouragingly, primitives of this sort do occasionally arise.
 For example, a recent revision of OCaml added support for  generalized algebraic data types (GADTs), based on research on guarded recursive datatype constructors \cite{XiCheChe03}.
-Using GADTs, OCaml was able to move some of the \emph{ad hoc} machinery for typechecking operations that use format strings, like \texttt{sprintf}, out of the language and into a library (note, however, that syntactic machinery remains  built in).
+Using GADTs, OCaml was able to move some of the \emph{ad hoc} machinery for typechecking operations that use format strings, like \texttt{sprintf}, out of the language and into a library.
+However, syntactic machinery remains  built in.
 
@@
 Our aim in the work being proposed is to introduce primitive language constructs that take further steps down the second path just described.
-In particular, we plan to introduce the following primtives:
+In particular, we plan to introduce the following primitives, which reduce the need for syntactic dialects:
 % By supporting the primitives that we introduce, 1) Verse will be smaller than comparable languages like ML and Scala, and 2) dialect formation will be less frequently necessary.

### 2015-10-06 thesis-proposal/problem.tex
Updated proposal to remove metamodules

@@
 %\subsection{Why are there so many dialects?}
-{This calls for an investigation}: why is it that programmers and researchers are still so often unable to satisfyingly express the constructs that they need in libraries, as modes of use of the ``general-purpose'' primitives already available in major languages today, and instead see a need for new language dialects?
-
-Some of these dialects extend the semantics of the language that they are based on, but the more common situation is that the existing semantic primitives suffice and it is only the \emph{syntactic cost} of expressing a construct of interest using these primitives  that is too high.
+%{This calls for an investigation}: why is it that programmers and researchers are still so often unable to satisfyingly express the constructs that they need in libraries, as modes of use of the ``general-purpose'' primitives already available in major languages today, and instead see a need for new language dialects?
+
+Although some of these dialects extend the semantics of the language that they are based on, the more common situation appears to be that the existing semantic primitives suffice and it is only the \emph{syntactic cost} of expressing one or more constructs of interest using these primitives  that is too high.
 In response, library providers construct \emph{syntactic dialects} -- dialects that introduce only new derived syntactic forms.
@@
 %Our broad aim in the work being proposed is to introduce primitive language mechanisms that give library providers the ability to  express new syntactic expansions as well as new types and operators in a safe and modularly composable manner.
-Our aim in the work being proposed is to introduce primitive language constructs that take further steps down the second path just described.
-In particular, we plan to introduce the following primitives, which reduce the need for syntactic dialects:
+Our aim in the work being proposed is to introduce primitive language constructs that take further steps down the second path just described by reducing the need for syntactic dialects.
+In particular, we plan to introduce the following primitives:
 % By supporting the primitives that we introduce, 1) Verse will be smaller than comparable languages like ML and Scala, and 2) dialect formation will be less frequently necessary.
@@
 \begin{enumerate}
-\item \textbf{Typed syntax macros} (TSMs), introduced in Sec \ref{sec:syntax}, reduce the need to primitively build in derived concrete syntactic forms specific to library constructs (e.g.
+\item \textbf{Typed syntax macros} (TSMs), introduced in Sec.
+\ref{sec:tsms}, reduce the need to primitively build in derived concrete syntactic forms specific to library constructs (e.g.
 list syntax as in SML or XML syntax as in Scala and Ur/Web), by giving library providers static control over the parsing and expansion of delimited segments of textual syntax (at a specified type or parameterized family of types).
-We begin by showing explicitly-invoked TSMs, then show how to associate a privileged TSM with a type declaration and then rely on a local type inference scheme to invoke these implicitly.
-\item \textbf{Metamodules}, introduced in Sec.
+\item \textbf{Type-specific languages} (TSLs), introduced in Sec.
+\ref{sec:tsls}, further reduce syntactic cost by allowing library providers to associate a TSM with a type declaration and then rely on a local type inference scheme to invoke it implicitly.
+%\item \textbf{Metamodules}, introduced in Sec.
 \ref{sec:metamodules}, reduce the need to primitively build in the type structure of constructs like records (and variants thereof),  labeled sums and other interesting constructs that we will introduce later by giving library providers programmatic ``hooks'' directly into the semantics, which are specified as a \emph{type-directed translation semantics} targeting a small \emph{typed internal language} (introduced in Sec.
@@
 \end{enumerate} 
-Both TSMs and metamodules make extensive use of \emph{static code generation} (also called \emph{static} or \emph{compile-time metaprogramming}, hence the name \emph{metamodules}), meaning that the relevant rules in the static semantics of the language call for the evaluation of \emph{static functions} that generate static representations of expressions and types.
-%Library providers write these static functions using the Verse \emph{static language} (SL).
-The design we are proposing also has conceptual roots in earlier work on \emph{active libraries}, which similarly envisioned using compile-time computation to give library providers more control over aspects of the language and compilation process (but did not take a type theoretic approach) \cite{active-libraries-thesis}.
+As vehicles for this work, we plan to formally specify a series of small typed lambda calculi that capture each of the novel primitives that we introduce ``minimally''.
+For the sake of examples, we will also describe (but not formally specify) a ``full-scale'' functional language called Verse.\footnote{We distinguish Verse from Wyvern, which is the language referred to in prior publications about some of the work that we will describe, because Wyvern is a group effort evolving independently in some important ways.} Verse is based semantically on Standard ML, differing mainly in that it uses a local type inference scheme \cite{Pierce:2000:LTI:345099.345100} (like, for example, Scala) for reasons that we will return to in Sec.
+\ref{sec:tsls}.
+The reason we will not follow Standard ML \cite{mthm97-for-dart} in giving a complete formal specification of Verse is both to emphasize that the primitives we introduce can be considered for inclusion in a variety of language designs, and to avoid distracting the reader with specifications for ``orthogonal'' primitives that are already well-understood in the literature.
+%We anticipate that future full-scale language specifications will be able to combine the ideas  in the proposed work without trouble.
+%The purpose of the work being proposed is to serve as a reference for those interested in the new constructs we introduce, not to serve as a language specification.
+%We will give a brief overview of these languages are organized in Sec. \ref{sec:verse}.
+
+The primitives we propose are forms of \emph{static code generation} (also sometimes called \emph{static} or \emph{compile-time metaprogramming}), meaning that the relevant rules in the static semantics of the language call for the evaluation of \emph{static functions} that generate static representations of terms.
+Static functions are functions written in a restricted subset of the language (we will discuss the necessary restrictions in Sec.
+\ref{sec:tsms-limitations}).%Library providers write these static functions using the Verse \emph{static language} (SL).
+
+The design we are proposing also has conceptual roots in earlier work on \emph{active libraries}, which similarly envisioned using compile-time computation to give library providers more control over aspects of the language (but did not take a type theoretic approach) \cite{active-libraries-thesis}.
 %Maintaining a separation between the static (or ``compile-time'') phase and the dynamic (or ``run-time'') phase is an important facet of Verse's design.
@@
 %We will  also introduce a simple variant of each of these primitives that leverages Verse's support for local type inference to further reduce syntactic cost in certain common situations.
-
-As vehicles for this work, we plan to formally specify small typed lambda calculi that capture each of the novel primitives that we introduce ``minimally''.
-We will also describe (but not formally specify) a new ``full-scale'' functional language called Verse.\footnote{We distinguish Verse from Wyvern, which is the language referred to in prior publications about some of the work that we will describe, because Wyvern is a group effort evolving independently in some important ways.} The reason we will not follow Standard ML \cite{mthm97-for-dart} in giving a complete formal specification of Verse is both to emphasize that the primitives we introduce can be considered for inclusion in a variety of language designs, and to avoid distracting the reader with specifications for ``orthogonal'' primitives that are already well-understood in the literature.
-%We anticipate that future full-scale language specifications will be able to combine the ideas  in the proposed work without trouble.
-%The purpose of the work being proposed is to serve as a reference for those interested in the new constructs we introduce, not to serve as a language specification.
-We will give a brief overview covering how these languages are organized in Chap. \ref{sec:verse}.
 
@@
 We will avoid this somewhat confusing terminology.} Our main technical contributions will be in rigorously showing how to address these problems in a principled manner.
-In particular, syntactic conflicts will be impossible by construction and the semantics will validate code statically generated by TSMs and metamodules to maintain:
+In particular, syntactic conflicts will be impossible by construction and the semantics will validate code statically generated by TSMs and TSLs to maintain:
 \begin{itemize}
-\item a \emph{hygienic type discipline}, meaning that these constructs maintain type safety and that the validity of the code generated will not make any assumptions about the surrounding context; and
+\item a \emph{hygienic type discipline}, meaning that the language is type safe, that one can reason about the type of a well-typed expression without examining its expansion, and that the expansion does not make any assumptions about the names of variables in the surrounding context; and
 \item \emph{modular reasoning principles}, meaning that library providers will have the ability to reason about the constructs that they have defined in isolation, and clients will be able to use them safely in any combination, without the possibility of conflict.\footnote{This is not quite true -- simple naming conflicts can arise.
@@
 \end{itemize}
-We will make these notions completely precise as we continue.
+%We will make these notions completely precise as we continue.
 
@@
 A functional programming language can give library providers the ability to %meta\-pro\-gram\-matic\-ally 
-express new syntactic expansions and new types and operators atop a small type-theoretic internal language while maintaining a hygienic type discipline and modular reasoning principles.
+express new syntactic expansions while maintaining a hygienic type discipline and modular reasoning principles.
 %These  primitives are  expressive enough to subsume the need for a variety of primitives that are, or would need to be, built in to comparable contemporary languages.
@@
 Before we continue, it may be useful to explicitly acknowledge that completely eliminating the need for dialects would indeed be asking for too much: certain design decisions are fundamentally incompatible with others or require coordination across a language design.
-We aim only to decrease the need for dialects.% out a larger design space within a single language, Verse.%a subset of constructs that can be specified by a semantics of a certain ``shape'' specified by Verse (we will make this more specific later).
+We aim only to decrease the need for syntactic dialects in this work.
+% out a larger design space within a single language, Verse.%a subset of constructs that can be specified by a semantics of a certain ``shape'' specified by Verse (we will make this more specific later).
 %There is nothing ``universal'' about Verse.
@@
 It may also be useful to explicitly acknowledge that library providers could leverage the primitives we introduce   to define constructs that are in ``poor taste''.
-We  expect that in practice, Verse will come with a standard library defining a carefully curated collection of standard constructs, as well as guidelines for advanced users regarding when it would be sensible to use the mechanisms we introduce (following the example of languages that support operator overloading or type classes \cite{Hall:1996:TCH:227699.227700}, which also have the potential for such ``abuse'').
+We  expect that in practice, Verse will come with a standard library defining a carefully curated collection of standard constructs, as well as guidelines for advanced users regarding when it would be sensible to use the mechanisms we introduce (following the example of languages that support operator overloading or type classes \cite{Hall:1996:TCH:227699.227700}, which also have the potential for ``abuse'').
 For most programmers, using Verse should not be substantially different from using a language like ML or one of its dialects.%The vast majority of programmers should not use the primitives that we introduce directly.
@@
 Verse is designed to be used for programming tasks where SML, OCaml, Haskell or Scala would be used today, not for advanced theorem proving tasks.
-That said, we conjecture that the primitives we describe could be added to languages like Gallina (the ``external language'' of the Coq proof assistant  \cite{Coq:manual}) or to the program extraction mechanisms of proof assistants like Coq with  modifications, but do not plan to pursue this line of research in this dissertation.
+That said, we conjecture that the primitives we describe could be added to languages like Gallina (the ``external language'' of the Coq proof assistant  \cite{Coq:manual}) with  modifications, but do not plan to pursue this line of research in this dissertation.
 %Our interest is in making sure that the standard library is no more privileged than any other library.
@@
 
-\section{Language Overview}\label{sec:verse}
-Let us begin with a brief overview of how Verse and the minimal calculi that we will develop are organized and specified at a high level.
+% \section{Language Overview}\label{sec:verse}
+% Let us begin with a brief overview of how Verse and the minimal calculi that we will develop are organized and specified at a high level.
 We will assume that readers are familiar with the early chapters of a textbook like \emph{TAPL} \cite{tapl} or \emph{PFPL} \cite{pfpl}.%Verse consists of a \emph{module language} atop a \emph{core language}.
 
-\subsection{Core Language}
-The Verse core language -- the language of types and expressions -- will be the focus of our efforts.
+% \subsection{Core Language}
+% The Verse core language -- the language of types and expressions -- will be the focus of our efforts.
 Figure \ref{fig:overview} gives a diagrammatic overview of how the core language is organized.
@@
 % The SL is notionally analagous to the language that the first stage itself is written in.
-We summarize the main judgements relevant to the IL, EL and SL below. Readers who prefer to see examples first can skim these sections for now, returning as needed.
-
-\begin{figure}
-\includegraphics[scale=0.5,clip=true,trim=0 120 0 120]{overview.pdf}\vspace{-10px}
-\caption{A diagrammatic overview of the core language.}
-\vspace{5px}
-\label{fig:overview}
-\end{figure}
-
-\subsubsection{Internal Language}
-Programs ultimately evaluate as \emph{internal expressions}, $\iota$.
+% We summarize the main judgements relevant to the IL, EL and SL below. Readers who prefer to see examples first can skim these sections for now, returning as needed.
+
+% \begin{figure}
+% \includegraphics[scale=0.5,clip=true,trim=0 120 0 120]{overview.pdf}\vspace{-10px}
+% \caption{A diagrammatic overview of the core language.}
+% \vspace{5px}
+% \label{fig:overview}
+% \end{figure}
+
+% \subsubsection{Internal Language}
+% Programs ultimately evaluate as \emph{internal expressions}, $\iota$.
 Internal expressions are classified by \emph{internal types}, $\tau$, so the internal language forms a standard typed lambda calculus.
@@
 The main judgements in the static semantics of the IL take the following familiar form (omitting contexts for simplicity throughout this section):
-\\[1ex]
-$
-\begin{array}{ll}
-\textbf{Judgement Form} & \textbf{Pronunciation}\\
-\vdash \tau~\mathtt{itype} & \text{Internal type $\tau$ is valid.}\\
-\vdash \iota : \tau & \text{Internal expression $\iota$ is assigned internal type $\tau$.}
-\end{array}
-$\\
-
-\noindent
-The dynamic semantics are specified also in the standard manner as a transition system with  judgements of the following form:
-\\[1ex]
-$
-\begin{array}{ll}
-\textbf{Judgement Form} & \textbf{Pronunciation}\\
-\iota \mapsto \iota' & \text{Internal expression $\iota$ transitions to $\iota'$.}\\
-\iota~\mathtt{val} & \text{Internal expression $\iota$ is a value.}
-\end{array}
-$
-\\[1ex]
-The iterated transition judgement $\iota \mapsto^{*} \iota'$ is the reflexive, transitive closure of the transition judgement, and the evaluation judgement $\iota \Downarrow \iota'$ is derivable iff $\iota \mapsto^{*} \iota'$ and $\iota'~\mathtt{val}$.
-
-Note that features like state, exceptions, basic concurrency primitives, scalars, arrays, I/O primitives and others characteristic of a first-stage compiler intermediate language would also be included in the IL in practice, and for several of these, this would affect the shape of the internal semantics.
+% \\[1ex]
+% $
+% \begin{array}{ll}
+% \textbf{Judgement Form} & \textbf{Pronunciation}\\
+% \vdash \tau~\mathtt{itype} & \text{Internal type $\tau$ is valid.}\\
+% \vdash \iota : \tau & \text{Internal expression $\iota$ is assigned internal type $\tau$.}
+% \end{array}
+% $\\
+
+% \noindent
+% The dynamic semantics are specified also in the standard manner as a transition system with  judgements of the following form:
+% \\[1ex]
+% $
+% \begin{array}{ll}
+% \textbf{Judgement Form} & \textbf{Pronunciation}\\
+% \iota \mapsto \iota' & \text{Internal expression $\iota$ transitions to $\iota'$.}\\
+% \iota~\mathtt{val} & \text{Internal expression $\iota$ is a value.}
+% \end{array}
+% $
+% \\[1ex]
+% The iterated transition judgement $\iota \mapsto^{*} \iota'$ is the reflexive, transitive closure of the transition judgement, and the evaluation judgement $\iota \Downarrow \iota'$ is derivable iff $\iota \mapsto^{*} \iota'$ and $\iota'~\mathtt{val}$.
+
+% Note that features like state, exceptions, basic concurrency primitives, scalars, arrays, I/O primitives and others characteristic of a first-stage compiler intermediate language would also be included in the IL in practice, and for several of these, this would affect the shape of the internal semantics.
 However, our design is largely insensitive to such details -- the only strict requirements are that the IL be type safe and support parametric type abstraction.
@@
 
-\subsubsection{External Language}
-The external language (EL) supports a richer syntax and type structure atop the IL.
+% \subsubsection{External Language}
+% The external language (EL) supports a richer syntax and type structure atop the IL.
 Most programmers will interact exclusively with the EL, so throughout this work, the words ``expression'' and ``type'' used without qualification refer to external expressions and types, respectively.
 The central judgements in the specification of the EL take the following form  (again omitting various contexts, which we will introduce progressively):
-\\[1ex]
-$\begin{array}{ll}
-\textbf{Judgement Form} & \textbf{Pronunciation}\\
-\vdash \sigma~\mathtt{type} \leadsto \tau & \text{Static expression $\sigma$ is a type with translation $\tau$.}\\
-\vdash e \Rightarrow \sigma \leadsto \iota & \text{Expression $e$ synthesizes type $\sigma$ and has translation $\iota$.}\\
-\vdash e \Leftarrow \sigma \leadsto \iota & \text{Expression $e$ analyzes against type $\sigma$ and has translation $\iota$.}
-\end{array}
-$\\[1ex]The main points to note here are that:
-\begin{itemize}
-\item External types are distinct from internal types (thus distinguishing this style of specification from the Harper-Stone elaboration semantics for Standard ML \cite{Harper00atype-theoretic}).
-\item The expression typing judgements are \emph{bidirectional}, i.e.
+% \\[1ex]
+% $\begin{array}{ll}
+% \textbf{Judgement Form} & \textbf{Pronunciation}\\
+% \vdash \sigma~\mathtt{type} \leadsto \tau & \text{Static expression $\sigma$ is a type with translation $\tau$.}\\
+% \vdash e \Rightarrow \sigma \leadsto \iota & \text{Expression $e$ synthesizes type $\sigma$ and has translation $\iota$.}\\
+% \vdash e \Leftarrow \sigma \leadsto \iota & \text{Expression $e$ analyzes against type $\sigma$ and has translation $\iota$.}
+% \end{array}
+% $\\[1ex]The main points to note here are that:
+% \begin{itemize}
+% \item External types are distinct from internal types (thus distinguishing this style of specification from the Harper-Stone elaboration semantics for Standard ML \cite{Harper00atype-theoretic}).
+% \item The expression typing judgements are \emph{bidirectional}, i.e.
 we make a judgemental distinction between \emph{type synthesis} (the type is an ``output'') and \emph{type analysis} (the type is an ``input'') \cite{Pierce:2000:LTI:345099.345100}.
@@
 %, primarily because it is not technically compatible with the primitives we will introduce (but also because we believe that compilers for languages that support only local type inference generate clearer error messages).
-\item  The dynamic behavior of an external expression is determined directly by its translation to the IL, i.e. there is no separate transition system governing the dynamics of external expressions.
-
-Proving type safety amounts to proving that the translation of every external expression of type $\sigma$ is an internal expression of type $\tau$, where $\tau$ is the translation of $\sigma$ (and thus, by type safety of the IL, evaluation does not ``go wrong'').
+% \item  The dynamic behavior of an external expression is determined directly by its translation to the IL, i.e. there is no separate transition system governing the dynamics of external expressions.
+
+% Proving type safety amounts to proving that the translation of every external expression of type $\sigma$ is an internal expression of type $\tau$, where $\tau$ is the translation of $\sigma$ (and thus, by type safety of the IL, evaluation does not ``go wrong'').
 This corresponds notionally to the idea of \emph{type-preserving compilation} in type-directed compilers \cite{tarditi+:til-OLD}.
-\end{itemize}
-
-\subsubsection{Static Language}
-Finally, the static language plays an important role in the semantics of the novel primitives that are the topic of this proposal.
+% \end{itemize}
+
+% \subsubsection{Static Language}
+% Finally, the static language plays an important role in the semantics of the novel primitives that are the topic of this proposal.
 The static language is simply another typed lambda calculus consisting of \emph{static expressions}, $\sigma$, classified by \emph{sorts}\footnote{We use the word ``sort'' to avoid confusion with the phrase ``static type''.
 Our use of the  word ``sort'' should be considered distinct from the various other uses of this word found in the literature on programming languages.}, $\uppi$, according to the following judgements (the ``static statics''):
-\\[1ex]
-$
-\begin{array}{ll}
-\textbf{Judgement Form} & \textbf{Pronunciation}\\
-\vdash \uppi~\mathtt{sort} & \text{Sort $\uppi$ is valid.}\\
-\vdash \sigma :: \uppi & \text{Static expression $\sigma$ has sort $\uppi$.}
-\end{array}
-$\\
-
-\noindent
-The ``static dynamics'' are specified as a transition system with judgements of the following basic form (though again omitting some necessary contexts):
-\\[1ex]
-$
-\begin{array}{ll}
-\textbf{Judgement Form} & \textbf{Pronunciation}\\
-\sigma \mapsto \sigma' & \text{Static expression $\sigma$ transitions to $\sigma'$.}\\
-\sigma~\mathtt{sval} & \text{Static expression $\sigma$ is a static value.}\\
-\sigma~\mathtt{styerr} & \text{Static expression $\sigma$ raises an external type error.}
-\end{array}
-$
-\\[1ex]
-The iterated static transition judgement $\sigma \mapsto^{*} \sigma'$ is the reflexive, transitive closure of the static transition judgement, and the static evaluation judgement $\sigma \Downarrow \sigma'$ is derivable iff $\sigma \mapsto^{*} \sigma'$ and $\sigma'~\mathtt{sval}$.
-
-As suggested by the form of the external type translation judgement above, external types are static expressions.
+% \\[1ex]
+% $
+% \begin{array}{ll}
+% \textbf{Judgement Form} & \textbf{Pronunciation}\\
+% \vdash \uppi~\mathtt{sort} & \text{Sort $\uppi$ is valid.}\\
+% \vdash \sigma :: \uppi & \text{Static expression $\sigma$ has sort $\uppi$.}
+% \end{array}
+% $\\
+
+% \noindent
+% The ``static dynamics'' are specified as a transition system with judgements of the following basic form (though again omitting some necessary contexts):
+% \\[1ex]
+% $
+% \begin{array}{ll}
+% \textbf{Judgement Form} & \textbf{Pronunciation}\\
+% \sigma \mapsto \sigma' & \text{Static expression $\sigma$ transitions to $\sigma'$.}\\
+% \sigma~\mathtt{sval} & \text{Static expression $\sigma$ is a static value.}\\
+% \sigma~\mathtt{styerr} & \text{Static expression $\sigma$ raises an external type error.}
+% \end{array}
+% $
+% \\[1ex]
+% The iterated static transition judgement $\sigma \mapsto^{*} \sigma'$ is the reflexive, transitive closure of the static transition judgement, and the static evaluation judgement $\sigma \Downarrow \sigma'$ is derivable iff $\sigma \mapsto^{*} \sigma'$ and $\sigma'~\mathtt{sval}$.
+
+% As suggested by the form of the external type translation judgement above, external types are static expressions.
 More specifically, external types are static values of a primitive sort $\mathsf{Ty}$.
@@
 \ref{sec:metamodules}.
-%One might notice the similarities between the specifications of the SL and the IL.
+% %One might notice the similarities between the specifications of the SL and the IL.
 In fact, the SL is a conservative extension of our IL.
@@
 
-% The reason why we refer to the syntactic sort $\sigma$ as a ``static expression'' will become more clear later.
-
-%The main novelty of our work is that the Verse EL builds in very little syntax or type structure \emph{a priori}.
+% % The reason why we refer to the syntactic sort $\sigma$ as a ``static expression'' will become more clear later.
+
+% %The main novelty of our work is that the Verse EL builds in very little syntax or type structure \emph{a priori}.
 %Instead, \textbf{metamodules} and \textbf{typed syntax macros} (TSMs) are expressive enough to render this unnecessary.
 Although metamodules are in some sense prior to TSMs, we will describe TSMs first, temporarily assuming that the EL does build in standard type structure, for pedagogical purposes.
-\subsection{Module Language}
-
-The Verse module system is taken directly (up to small syntactic changes) from Standard ML, with which we assume a working familiarity for the purposes of this proposal \cite{harper1997programming,MacQueen:1984:MSM:800055.802036} (a related module system, e.g.
+% \subsection{Module Language}
+
+% The Verse module system is taken directly (up to small syntactic changes) from Standard ML, with which we assume a working familiarity for the purposes of this proposal \cite{harper1997programming,MacQueen:1984:MSM:800055.802036} (a related module system, e.g.
 OCaml's, would work just as well for our purposes).
@@
 
-Formally, our approach will be to first specify our contributions assuming a language without an ML-style module system, then make modifications to this specification to make it compatible with the presence of a module system.
+% Formally, our approach will be to first specify our contributions assuming a language without an ML-style module system, then make modifications to this specification to make it compatible with the presence of a module system.
 Because it has been thoroughly studied in the literature, we will defer to prior work both here and in the dissertation for many formal details.
@@
 
-\section{Modularly Programmable Textual Syntax}\label{sec:syntax}
-To begin, let us assume a standard, ML-like semantics for both the EL and SL and consider the EL's concrete syntax.
-Verse, like most major contemporary languages, specifies a textual concrete syntax.\footnote{Although Wyvern specified a layout-sensitive concrete syntax, to avoid unnecessary distractions, we will describe a more conventional layout-insensitive concrete syntax for Verse.} %We have chosen to specify a layout-sensitive textual concrete syntax (i.e.
+\section{Background}\label{sec:background}
+%To begin, let us assume a standard, ML-like semantics for both the EL and SL and consider the EL's concrete syntax. 
+Verse, like most contemporary languages, specifies a textual concrete syntax.\footnote{Although Wyvern specified a layout-sensitive concrete syntax, to avoid unnecessary distractions, we will describe a more conventional layout-insensitive concrete syntax for Verse.} %We have chosen to specify a layout-sensitive textual concrete syntax (i.e.
 newlines and indentation are not ignored).
@@
 %For constructs that have an obvious analog in ML, we will omit a detailed explanation.
-Because the purpose of this syntax is to serve as the programmer-facing user interface of the language, it is common practice to build in  derived syntactic forms (colloquially, \emph{syntactic sugar}) that capture common idioms more concisely or naturally.
+Because the purpose of this syntax is to serve as the programmer-facing user interface of the language, it is common practice to build in  derived syntactic forms (colloquially, \emph{syntactic sugar}) that capture common idioms more concisely or ``naturally''.
 % (i.e.
@@
 We will begin in Sec.
-\ref{sec:examples} by detailing another example for which such a mechanism would be useful: regular expression patterns expressed using abstract data types.
+\ref{sec:examples} by detailing another example for which such a mechanism would be useful: regular expression (regex) patterns expressed using abstract data types.
+We will refer to this example throughout the proposal.
 In Sec.
-\ref{sec:syntax-existing}, we then demonstrate that the usual approach of using dynamic string parsing to introduce patterns is not ideal.
+\ref{sec:syntax-existing}, we discuss how the usual approach of using dynamic string parsing to introduce regex patterns is not ideal.
 We also survey existing alternatives to dynamic string parsing, finding that they involve an unacceptable loss of modularity and other undesirable trade-offs.
-In Sec.
-\ref{sec:syntax-contributions}, we outline our proposed alternatives -- \emph{typed syntax macros} (TSMs) and the related \emph{type-specific languages} (TSLs) -- and discuss how they resolve these issues.
-We also give an overview of how TSMs are specified in terms of the judgement forms that were introduced in the previous section.
-We conclude in Sec.
-\ref{sec:syntax-timeline} with a concrete timeline for the remaining work.
+In Secs.
+\ref{sec:tsms} and \ref{sec:tsls}, we introduce our proposed alternatives -- \emph{typed syntax macros} (TSMs) and the related \emph{type-specific languages} (TSLs) -- and discuss how they resolve these issues (as well as some limitations that they have).
+We  also give an overview of how TSMs are formally specified.
+We give a concrete timeline for the remaining work in Sec.
+\ref{sec:syntax-timeline}, and conclude in Sec.
+\ref{sec:conclusion}.
 
@@
 Let us begin by taking the perspective of a regular expression library provider.
-We assume the reader has some familiarity with regular expressions \cite{Thompson:1968:PTR:363347.363387}.
+We assume the reader has some familiarity with regular expressions \cite{Thompson:1968:PTR:363347.363387} and Standard ML \cite{harper1997programming}.
 We will discuss a standard variant of regular expressions that supports marking \emph{captured groups} (with parentheses in the concrete syntax) to make certain examples more interesting.
 
-\paragraph{Abstract Syntax} The abstract syntax of {patterns}, $p$, over strings, $s$, is specified as below:\[p ::= \textbf{empty} ~|~ \textbf{str}(s) ~|~ \textbf{seq}(p; p) ~|~ \textbf{or}(p; p) ~|~ \textbf{star}(p) ~|~ \textbf{group}(p)\]
-One way to express this abstract syntax is by defining a recursive sum type \cite{pfpl}.
-Verse supports these as datatypes, which are comparable to datatypes in ML (we plan to show how the type structure  of datatypes can in fact be expressed in libraries later):
-
-\begin{lstlisting}[numbers=none]
-datatype Pattern {
-    Empty | Str of string | Seq of Pattern * Pattern 
-    | Or of Pattern * Pattern | Star of Pattern | Group of Pattern
+\paragraph{Abstract Syntax} The abstract syntax of {patterns}, $r$, over strings, $s$, is specified as below:\[r ::= \textbf{empty} ~|~ \textbf{str}(s) ~|~ \textbf{seq}(r; r) ~|~ \textbf{or}(r; r) ~|~ \textbf{star}(r) ~|~ \textbf{group}(r)\]
+One way to express this abstract syntax is by defining a recursive sum type \cite{pfpl}. Verse supports these as datatypes:
+
+\begin{lstlisting}[numbers=none]
+datatype Rx {
+    Empty | Str of string | Seq of Rx * Rx 
+    | Or of Rx * Rx | Star of Rx | Group of Rx
 }
@@
 First, regular expression patterns are usually identified up to their reduction to a normal form.
-For example, $\textbf{seq}(\textbf{empty}, p)$ has normal form $p$.
+For example, $\textbf{seq}(\textbf{empty}, r)$ has normal form $r$.
 It might be useful for patterns with the same normal form to be  indistinguishable from the perspective of client code.
@@
 Indeed, there may be many ways to represent regular expression patterns, each with different performance trade-offs.
-For these reasons, a better approach in Verse, as in ML, is to abstract over the choice of representation using  the module system's support for generative type abstraction.
+For these reasons, a better approach in Verse, as in ML, is to abstract over the choice of representation using  the module system's support for type abstraction.
 In particular, we can define the following \emph{module signature}, where the type of patterns, \lstinline{t}, is held abstract:
@@
 \begin{lstlisting}[deletekeywords={case},numbers=none]
-signature PATTERN = sig {
+signature RX = sig {
   type t
@@
   val case : (
-    t -> 
-    'a -> (* Empty case *)
-    (string -> 'a) -> (* Str case *)
-    (t * t -> 'a) -> (* Seq case *)
-    (t * t -> 'a) -> (* Or case *)
-    (t -> 'a) -> (* Star case *)
-    (t -> 'a) -> (* Group case *)
-    'a)
+    t -> {
+    	Empty : 'a,
+    	Str : string -> 'a,
+    	Seq : t * t -> 'a,
+    	Or : t * t -> 'a,
+    	Star : t -> 'a,
+    	Group : t -> 'a
+    } -> 'a
 }
 \end{lstlisting}
- Clients of any module \lstinline{P} that has been opaquely ascribed \lstinline{PATTERN}, written \lstinline{P :> PATTERN}, manipulate patterns as values of the type \verb|P.t| using the interface described by this signature.
-The identity of this type is held abstract outside the module during typechecking (i.e.
+ Clients of any module \lstinline{R} that has been sealed against \lstinline{RX}, written \lstinline{R :> RX}, manipulate patterns as values of the type \verb|R.t| using the interface described by this signature.
+The identity of the type \lstinline{R.t} is held abstract outside the module during typechecking (i.e.
 it acts as a newly generated type).
@@
 \paragraph{Concrete Syntax} The abstract syntax of patterns is too verbose to be practical  in all but the most trivial examples, so programmers conventionally write patterns using a more concise concrete syntax.
-For example, the concrete syntax \lstinline{A|T|G|C} corresponds to the following much more verbose pattern expression:
+For example, the concrete syntax \lstinline{A|T|G|C} corresponds to the following much more verbose pattern expression (assuming some module \lstinline{R : RX} is in scope here and in the remainder of the examples):
 \begin{lstlisting}[numbers=none,mathescape=|]
-P.Or(P.Str "SSTRAESTR", P.Or(P.Str "SSTRTESTR", P.Or(P.Str "SSTRGESTR", P.Str "SSTRCESTR")))
+R.Or(R.Str "SSTRAESTR", R.Or(R.Str "SSTRTESTR", R.Or(R.Str "SSTRGESTR", R.Str "SSTRCESTR")))
 \end{lstlisting}
@@
 To expose this more concise concrete syntax for regular expression patterns to clients, the most common approach is to provide a function that parses strings to produce patterns.
-Because, as just mentioned, there  may be many implementations of the \lstinline{PATTERN} signature, the usual approach is to define a parameterized module (a.k.a.
+Because, as just mentioned, there  may be many implementations of the \lstinline{RX} signature, the usual approach is to define a parameterized module (a.k.a.
 a \emph{functor} in SML) defining utility functions like this abstractly:
@@
 \begin{lstlisting}[numbers=none]
-module PatternUtil(P : PATTERN) => mod {
-  fun parse(s : string) : P.t => (* ... pattern parser here ... *)
+module RXUtil(R : RX) => mod {
+  fun parse(s : string) : R.t => (* ... regex parser here ... *)
 }
 \end{lstlisting}
-This allows a client of any module \lstinline{P : PATTERN} to use the following definitions:
-\begin{lstlisting}[numbers=none]
-let module PU = PatternUtil(P)
-let val pattern = PU.parse
+This allows a client of any module \lstinline{R : RX} to use the following definitions:
+\begin{lstlisting}[numbers=none]
+let module RUtil = RXUtil(R)
+let val rxparse = RUtil.parse
 \end{lstlisting}
@@
 \begin{lstlisting}[numbers=none]
-pattern "SSTRA|T|G|CESTR"
+rxparse "SSTRA|T|G|CESTR"
 \end{lstlisting}
@@
 \begin{lstlisting}[numbers=none,mathescape=|]
-let val ssn = pattern "SSTR\d\d\d-\d\d-\d\d\d\dESTR"
-\end{lstlisting}
-When compiling an expression like this, the client would see an error message like \verb|error: illegal escape character|.\footnote{This is the error message that \texttt{javac} produces.
-When compiling an analagous expression using SML of New Jersey (SML/NJ), we encounter a more confusing error message: \texttt{Error: unclosed string}.} In a small lab study, we observed that this class of error was common, and often confused even experienced programmers if they had not used regular expressions recently \cite{Omar:2012:ACC:2337223.2337324}.
-The standard workaround has higher syntactic cost -- we must double all backslashes:
-\begin{lstlisting}[numbers=none]
-let val ssn = pattern "SSTR\\d\\d\\d-\\d\\d-\\d\\d\\d\\dESTR"
+let val ssn = rxparse "SSTR\d\d\d-\d\d-\d\d\d\dESTR"
+\end{lstlisting}
+When compiling an expression like this, the client would see an error message like \verb|error: illegal escape character|\footnote{This is the error message that \texttt{javac} produces.
+When compiling an analagous expression using SML of New Jersey (SML/NJ), we encounter a more confusing error message: \texttt{Error: unclosed string}.}, because \verb|\d| is not a valid string escape sequence.
+In a small lab study, we observed that this class of error often confused even experienced programmers if they had not used regular expressions recently \cite{Omar:2012:ACC:2337223.2337324}.
+One workaround has higher syntactic cost -- we must double all backslashes:
+\begin{lstlisting}[numbers=none]
+let val ssn = rxparse "SSTR\\d\\d\\d-\\d\\d-\\d\\d\\d\\dESTR"
 \end{lstlisting}
@@
 \begin{lstlisting}[numbers=none]
-let val ssn = pattern {rx|SSTR\d\d\d-\d\d-\d\d\d\dESTR|rx}
+let val ssn = rxparse {rx|SSTR\d\d\d-\d\d-\d\d\d\dESTR|rx}
 \end{lstlisting}
@@
 Patterns constructed compositionally cannot easily benefit from this technique.
-For example, consider this function from strings to patterns:
+For example, consider the following function from strings to patterns:
 \begin{lstlisting}[numbers=none]
 fun example(name : string) => 
-  P.Seq(P.Str(name), P.Seq(pattern "SSTR: ESTR", ssn)) (* ssn as above *)
+  R.Seq(R.Str(name), R.Seq(rxparse "SSTR: ESTR", ssn)) (* ssn as above *)
 \end{lstlisting}
 %We needed to use both dynamic string parsing and explicit applications of pattern constructors to achieve the intended semantics. 
-Had we built derived syntax for regular expression patterns into the language primitively (following Unix conventions of using forward slashes as delimiters), we could use equivalently use splicing syntax:
-\begin{lstlisting}[numbers=none]
-fun example_shorter(name : string) => /@name: %ssn/
-\end{lstlisting}
-Here, an identifier (or parenthesized expression, not shown) prefixed with an \lstinline{@} is a spliced string, and one prefixed with a \lstinline{%} is a spliced pattern.
+Had we built derived syntax for regular expression patterns into the language primitively (following Unix conventions of using forward slashes as delimiters), we could have used \emph{splicing syntax}:
+\begin{lstlisting}[numbers=none]
+fun example_shorter(name : string) => /SURL@EURLnameSURL: %EURLssn/
+\end{lstlisting}
+An identifier (or parenthesized expression, not shown) prefixed with an \lstinline{@} is a spliced string, and one prefixed with a \lstinline{%} is a spliced pattern.
 
@@
 fun example_bad(name : string) => 
-  pattern (name ^ {rx|SSTR: \d\d\d-\d\d-\d\d\d\dESTR|rx})
+  rxparse (name ^ {rx|SSTR: \d\d\d-\d\d-\d\d\d\dESTR|rx})
 \end{lstlisting}
@@
 
-Both \lstinline{example} and \lstinline{example_bad} have the same type and behave identically at many inputs, particularly those that would be expected during typical executions of the program (i.e.
+Both \lstinline{example} and \lstinline{example_bad} have the same type and behave identically at many inputs, particularly ``typical'' inputs (i.e.
 alphabetic names).
@@
 case(moon_phase) {
-    Full => pattern "SSTR(GCESTR" (* malformedness not statically detected *)
+    Full => rxparse "SSTR(GCESTR" (* malformedness not statically detected *)
   | _ => (* ... *)
@@
 This is an undecidable verification problem in general and can be difficult to even partially automate.
-In this example, the verification procedure would first need to be able to establish that the variable \lstinline{pattern} is equal to the parse function \lstinline{PU.parse}.
+In this example, the verification procedure would first need to be able to establish that the variable \lstinline{rxparse} is equal to the parse function \lstinline{RUtil.parse}.
 If the string argument had not been written literally but rather computed, e.g.
 as \lstinline{"SSTR(GESTR" ^ "SSTRCESTR"} where \lstinline{^} is the string concatenation function applied in infix style, it would also need to be able to establish that this expression is equivalent to the string \lstinline{"SSTR(GCESTR"}.
-For patterns that are dynamically constructed based on input to a function, the problem is not solved by   evaluating the expression statically (or, more generally, in some earlier ``stage'' of evaluation) \cite{Jones:Gomard:Sestoft:93:PartialEvaluation}.
-
-Of course, asking the client to provide a proof of well-formedness would completely defeat the purpose -- lowering syntactic cost.
-
-In a language that primitively supported derived pattern syntax, the parsing would occur at compile-time and so malformed patterns would produce a compile-time error.
+For patterns that are dynamically constructed based on input to a function, evaluating the expression statically (or, more generally, in some earlier ``stage'' of evaluation \cite{Jones:Gomard:Sestoft:93:PartialEvaluation}) also does not suffice.
+
+Of course, asking the client to provide a proof of well-formedness would defeat the purpose of lowering syntactic cost.
+
+In contrast, were our language to primitively support  derived pattern syntax, pattern parsing would occur at compile-time and so malformed patterns would produce a compile-time error.
 
@@
 \begin{lstlisting}[numbers=none]
-map exmpl_list (fn s => rx_match (pattern "SSTRA|T|G|CESTR") s)
+map exmpl_list (fn s => rxmatch (rxparse "SSTRA|T|G|CESTR") s)
 \end{lstlisting}
@@
 
-In a language that primitively supported derived pattern syntax, the expansion is computed at compile-time and so there is no dynamic cost.
+In contrast, were our language to primitively support derived pattern syntax, the expansion would be computed at compile-time and incur no dynamic cost.
 \end{enumerate}
@@
 Unfortunately, a single equation is not enough to allow us to express pattern syntax following the usual conventions.
-For example, there is no way to handle sequences of characters (they can only be treated as variables) or escape characters (there is no way to define disjunctive syntax) using a system like Coq's.
+For example, a system like Coq's cannot handle escape characters, because there is no way to programmatically examine a form when generating its expansion.
 
@@
 For example, many are based on context-free grammars, e.g.
-Sugar* \cite{erdweg2013framework}, Camlp4 \cite{ocaml-manual} and several others.
-Other systems give library providers direct programmatic access to the parse stream, like Common Lisp's \emph{reader macros} \cite{steele1990common} (which are distinct from its term-rewriting macros, described below) and Racket's preprocessor \cite{Flatt:2012:CLR:2063176.2063195}.
-All of these would allow us to add pattern syntax into our core language's grammar, perhaps following Unix conventions like this:
-\begin{lstlisting}[numbers=none]
-let val ssn = /\d\d\d-\d\d-\d\d\d\d/
-\end{lstlisting}
-
-As discussed above, we can define \emph{splicing syntax} that allows us to write string and pattern expressions inline (distinguished by prefixes \lstinline{@} and \lstinline{%}, respectively). 
-For example, we can write \lstinline{example} equivalently, but at lower syntactic cost (cf. \lstinline{example_shorter}).
+Sugar* \cite{erdweg2013framework} and Camlp4 \cite{ocaml-manual} (amongst many others).
+Other systems give library providers direct programmatic access to the parse stream, like Common Lisp's \emph{reader macros} \cite{steele1990common} (which are distinct from its term-rewriting macros, described in Sec.
+\ref{sec:term-rewriting} below) and Racket's preprocessor \cite{Flatt:2012:CLR:2063176.2063195}.
+All of these would allow us to add pattern syntax into our language's grammar, perhaps following Unix conventions and supporting splicing syntax as described above:
+\begin{lstlisting}[numbers=none]
+let val ssn = /SURL\d\d\d-\d\d-\d\d\d\dEURL/
+fun example_shorter(name : string) => /SURL@EURLnameSURL: %EURLssn/
+\end{lstlisting}
 %The body of this function elaborates to the body of \lstinline{example_fixed} as shown above. 
@@
 
-We sidestep the problems of dynamic string parsing described above  if we directly extend the syntax of our language using any of these approaches.
+We sidestep the problems of dynamic string parsing described above  when we directly extend the syntax of our language using any of these systems.
 Unfortunately, direct syntax extension introduces serious new problems.
@@
 
-In response to this problem, Schwerdfeger and Van Wyk developed a modular analysis that accepts only grammar extensions that specify a unique starting token and obey subtle constraints on  the follow sets of base language non-terminals \cite{conf/pldi/SchwerdfegerW09}.
-Extensions that satisfy these criteria can be used together in any combination without the possibility of syntactic conflict.
-However, the most natural starting tokens like \lstinline{pattern} cannot be guaranteed to be unique.
+In response to this problem, Schwerdfeger and Van Wyk developed a modular analysis that accepts only context-free grammar extensions that begin with an identifying starting token and obey certain constraints on  the follow sets of base language's non-terminals \cite{conf/pldi/SchwerdfegerW09}.
+Extensions that specify distinct starting tokens and that satisfy these constraints can be used together in any combination without the possibility of syntactic conflict.
+However, the most natural starting tokens like \lstinline{rx} cannot be guaranteed to be unique.
 To address this problem, programmers must agree on a convention for defining ``globally unique identifiers'', e.g.
-the common URI convention used on the web and by the Java ecosystem.
-This forces us to use a more verbose token like \lstinline{edu_cmu_verse_rx_pattern}.
-There is no simple way for clients of our extension to define scoped abbreviations for starting tokens because this mechanism operates purely at the level of syntactic forms.
-
-Putting this aside, we must also consider the question of how newly introduced forms expand to forms in our base grammar.
-In our example, this is tricky because we have defined a modular encoding of patterns -- which particular module should the expansion use?
-Clearly, simply assuming that some module identified as \lstinline{P} matching \lstinline{PATTERN} is in scope is a brittle solution.
-In fact, we should expect that the system actively prevents such capture of specific variable names to ensure that variables (including module variables) can be freely renamed.
+the common URI convention used on the web and by the Java packaging system.
+However, this forces us to use a more verbose token like \lstinline{edu_cmu_verse_rx}.
+There is no simple way for clients of our extension to define scoped abbreviations for starting tokens because this mechanism operates purely at the level of the context-free grammar.
+
+Putting this aside, we must also consider another modularity-related question: which particular module should the expansion use?
+Clearly, simply assuming that some module identified as \lstinline{R} matching \lstinline{RX} is in scope is a brittle solution.
+In fact, we should expect that the system actively prevents such capture of specific variable names to ensure that variables (here, module variables) can be freely renamed.
 Such a \emph{hygiene discipline} is well-understood only when performing term-to-term rewriting (discussed below) or in simple language-integrated rewrite systems like those found in Coq.
-For mechanisms that operate strictly at the level of context-free grammars or the parse stream, this issue has not been adequately addressed.
-The library provider must simply be careful not to make assumptions about variable names and instead require that the client explicitly identify the module they intend to use as an ``argument'' in the new form:
-\begin{lstlisting}[numbers=none]
-let val ssn = edu_cmu_verse_rx_pattern P /\d\d\d-\d\d-\d\d\d\d/
+For mechanisms that operate strictly at the level of context-free grammars or the parse stream, it is not clear how one could address this issue.
+The onus is then on the library provider to make no assumptions about variable names and instead require that the client explicitly identify the module they intend to use as an ``argument'' within the newly introduced form:
+\begin{lstlisting}[numbers=none]
+let val ssn = edu_cmu_verse_rx R /SURL\d\d\d-\d\d-\d\d\d\dEURL/
 \end{lstlisting}
@@
 Early variants of this system suffered from the problem of unhygienic variable capture just described, but  later variants, notably in the Scheme dialect of LISP, brought support for enforcing hygiene \cite{Kohlbecker86a}.
-In languages with a stronger static type discipline, variants of macros that restrict rewriting to a particular type and perform the rewriting statically have also been studied \cite{Herman10:Theory,ganz2001macros} and integrated into languages, e.g.
+In languages with a richer static type discipline, variants of macros that restrict rewriting to a particular type and perform the rewriting statically have also been studied \cite{Herman10:Theory,ganz2001macros} and integrated into languages, e.g.
 MacroML \cite{ganz2001macros} and Scala \cite{ScalaMacros2013}.
 
-The most immediate problem with using these for our example is that we are not aware of any such statically-typed macro system that integrates cleanly with an ML-style module system, i.e.
-macros cannot be parameterized by modules.
+The most immediate problem with using these for our example is that we are not aware of any such statically-typed macro system that integrates cleanly with an ML-style module system.
+In other words, macros cannot be parameterized by modules.
 However, let us imagine such a macro system.
@@
 \begin{lstlisting}[numbers=none]
-let val ssn = pattern P {rx|SSTR\d\d\d-\d\d-\d\d\d\dESTR|rx}
-\end{lstlisting}
-
-The definition of \lstinline{pattern} might look like this:
-\begin{lstlisting}[numbers=none]
-macro pattern[Q : PATTERN](e) at Q.t {
-  static fun f(e :: Exp) :: Exp => case(e) {
-      StrLit(s) => (* pattern parser here *)
-    | BinOp(Caret, e1, e2) => `Q.Seq(Q.Str(%e1), %(f e2))`
-    | BinOp(Plus, e1, e2) => `Q.Seq(%(f e1), %(f e2))`
+let val ssn = rx R {rx|SSTR\d\d\d-\d\d-\d\d\d\dESTR|rx}
+\end{lstlisting}
+
+The definition of the macro \lstinline{rx} might look like this:
+\begin{lstlisting}
+macro rx[Q : RX](e) at Q.t {
+  static fun f(e : Exp) : Exp => case(e) {
+      StrLit(s) => (* regex parser here *)
+    | BinOp(Caret, e1, e2) => `SQTQ.Seq(Q.Str(%EQTe1SQT), %(EQTf e2SQT))EQT`
+    | BinOp(Plus, e1, e2) => `SQTQ.Seq(%(EQTf e1SQT), %(EQTf e2SQT))EQT`
     | _ => raise Error
@@
 
-Here, \lstinline{pattern} is a macro parameterized by a module satisfying the signature \lstinline{PATTERN} (we identify it as \lstinline{Q} to emphasize that there is nothing special about the identifier \lstinline{P} that we have been using) and taking a single argument, identified as \lstinline{e}.
-The macro specifies that the expansion it statically generates will be of type \lstinline{Q.t}.
-It is defined by a static function that examines the syntax tree of the provided argument (syntax trees are of a sort \lstinline{Exp} defined in the standard library; cf.
+Here, \lstinline{rx} is a macro parameterized by a module matching \lstinline{rx} (we identify it as \lstinline{Q} to emphasize that there is nothing special about the identifier \lstinline{R}) and taking a single argument, identified as \lstinline{e}.
+The macro specifies a type annotation, \lstinline{at Q.t}, which imposes the constraint that the expansion the macro statically generates must be of type \lstinline{Q.t} for the provided parameter \lstinline{Q}.
+This expansion is generated by a \emph{static function} that examines the syntax tree of the provided argument (syntax trees are of a type \lstinline{Exp} defined in the standard library; cf.
 SML/NJ's visible compiler \cite{SML/VisibleCompiler}).
-If it is a string literal, as in the example above, it statically parses the literal body to generate an expansion (the details of the parser would be entirely standard).
+If it is a string literal, as in the example above, it statically parses the literal body to generate an expansion (the details of the parser, elided on line 3, would be entirely standard).
 By parsing the string statically, we avoid the problems of dynamic string parsing for statically-known patterns.
@@
 For patterns that are constructed compositionally, we need to get more creative.
-For example, we might repurpose the infix operators that are normally used for other purposes to support string and pattern splicing as follows:
+For example, we might repurpose the infix operators that are normally used for other purposes to support string and pattern splicing, e.g.
+as follows:
 
 \begin{lstlisting}[numbers=none,escapechar=|]
-fun example_macro(name : string) => 
-  pattern P (name ^ "SSTR: ESTR" + ssn)
-\end{lstlisting}
-
-The logic implementing this can be seen above.
-We assume that there is derived syntax available for the sort \lstinline{Exp} where the \lstinline{%} prefix indicates a spliced expression, i.e.
-\emph{quasiquotation} as in Lisp \cite{Bawd99a} and Scala \cite{shabalin2013quasiquotes}.
-
-Having to creatively repurpose existing syntax in this way limits the effect a library provider can have on syntactic cost (particularly when it would be desirable to adopt conventions that are not consistent with the conventions adopted by the language).
+fun example_using_macro(name : string) => 
+  rx R (name ^ "SSTR: ESTR" + ssn)
+\end{lstlisting}
+
+The binary operator \lstinline{^} is repurposed to indicate a spliced string and \lstinline{+} is repurposed to indicate a spliced pattern.
+The logic for handling these forms can be seen above on lines 4 and 5, respectively.
+We assume that there is derived syntax available at the type \lstinline{Exp}, i.e.
+\emph{quasiquotation} syntax as in Lisp \cite{Bawd99a} and Scala \cite{shabalin2013quasiquotes}, here delimited by backticks and using the prefix \lstinline{%} to indicate a spliced value (i.e.
+unquote).
+
+Having to creatively repurpose existing forms in this way limits the effect a library provider can have on syntactic cost (particularly when it would be desirable to express conventions that are quite different from the conventions adopted by the language).
 It also can create confusion for readers expecting parenthesized expressions to behave in a consistent manner.
-However,  this approach is still preferable to direct syntax extension because it avoids many of the problems discussed above: there cannot be syntactic conflicts (because the syntax is not extended at all), we can define macro abbreviations because macros are integrated into the language, there is a hygiene discipline that guarantees that the expansion will not capture variables inadvertently, and by using a typed macro system, programmers need not examine the expansion to know what type a macro application must have.
+However,  this approach is preferable to direct syntax extension because it avoids many of the problems discussed above: there cannot be syntactic conflicts (because the syntax is not extended at all), we can define macro abbreviations because macros are integrated into the language, there is a hygiene discipline that guarantees that the expansion will not capture variables inadvertently, and by using a typed macro system, programmers need not examine the expansion to know what type the expansion produced by a macro must have.
 
@@
 
-\subsection{Contributions}\label{sec:syntax-contributions}
-Verse provides a new external primitive -- the \textbf{typed syntax macro} (TSM) -- that combines the syntactic flexibility of syntax extensions with the reasoning guarantees of typed macros.
-TSMs can be parameterized by modules, so they can be used to define syntax valid at any type defined by a module satisfying a specified signature.
-As we will discuss in Sec.
-\ref{sec:tsms} below, this addresses all of the problems brought up above, at moderate syntactic cost.
-
-To further decrease syntactic cost (which is, of course, the entire purpose of this effort), we go on in Sec.
-\ref{sec:tsls} to briefly introduce \textbf{type-specific languages} (TSLs).
-TSLs are TSMs associated directly with a type when it is generated.
-Verse uses local type inference to implicitly control TSL dispatch.
-
-Syntax defined by library providers using TSMs and TSLs comes at syntactic cost comparable to that of derived syntax built in primitively by the language designer or  using a na\"ive syntax extension tool like Camlp4, without causing dialect formation.
-As such, these \emph{ad hoc} approaches can, in large part, be discarded.
-That said, we examine some limitations of our approach in Sec.
-\ref{sec:tsms-limitations}.
-
-\subsubsection{Typed Syntax Macros (TSMs)}\label{sec:tsms}
+\section{Typed Syntax Macros (TSMs)}\label{sec:tsms}
+We will now introduce a new external primitive -- the \textbf{typed syntax macro} (TSM) -- that combines the syntactic flexibility of syntax extensions with the reasoning guarantees of typed macros.
+Like the term-rewriting macros just described, TSMs can be parameterized by modules, so they can be used to define syntax valid at any abstract type defined by a module satisfying a specified signature.
+As we will discuss in the remainder of this section, this addresses all of the problems brought up above, at moderate syntactic cost.
+
+\subsection{TSMs by Example}\label{sec:tsms-by-example}
 %A typed syntax macro is invoked by applying it to a \emph{delimited form}, which can contain  arbitrary syntax in its \emph{body}.  
-To introduce TSMs, let us consider the following concrete external expression:
-\begin{lstlisting}[numbers=none]
-pattern P /SURLA|T|G|CEURL/
-\end{lstlisting}
-Here, we apply a \emph{parameterized TSM}, \lstinline{pattern}, first to a module parameter, \lstinline{P}, then to a \emph{delimited form}, \lstinline{/SURLA|T|G|CEURL/}.
-Note that a number of alternative delimiters are also provided by Verse's concrete syntax and could equivalently be used.
+To introduce TSMs, consider the following concrete external expression:
+\begin{lstlisting}[numbers=none]
+rx P /SURLA|T|G|CEURL/
+\end{lstlisting}
+Here, we apply a \emph{parameterized TSM}, \lstinline{rx}, first to a module parameter, \lstinline{R}, then to a \emph{delimited form}, \lstinline{/SURLA|T|G|CEURL/}.
+A number of alternative delimiters are also available in Verse's concrete syntax and could equivalently have been used, e.g.
+\lstinline{<SURLA|T|G|CEURL>}.
 The TSM statically parses the \emph{body} of the provided delimited form, i.e.
-the characters between the delimiters (shown here in blue), and computes an \emph{expansion}, i.e.
-another external expression.
+the characters between the delimiters (indicated here in blue), and computes an \emph{expansion}, i.e.
+another expression.
 In this case, \lstinline{pattern} generates the following expansion (written concretely):
@@
 \begin{lstlisting}[numbers=none]
-P.Or(P.Str "SSTRAESTR", P.Or(P.Str "SSTRTESTR", P.Or(P.Str "SSTRGESTR", P.Str "SSTRCESTR")))
+R.Or(R.Str "SSTRAESTR", R.Or(R.Str "SSTRTESTR", R.Or(R.Str "SSTRGESTR", R.Str "SSTRCESTR")))
 \end{lstlisting}
@@
 %The original expression, above, is statically rewritten to this expression.
-The definition of \lstinline{pattern} looks like this:
-\begin{lstlisting}[numbers=none]
-syntax pattern(Q : PATTERN) at Q.t {
-  static fn (body :: Body) :: Exp => (* pattern parser here *)
+The definition of \lstinline{rx} looks like this:
+\begin{lstlisting}[numbers=none]
+syntax rx[Q : RX] at Q.t {
+  static fn (body : Body) : Exp => (* regex expression parser here *)
 }
 \end{lstlisting}
-This TSM definition first identifies the TSM as \lstinline{pattern}, then specifies a module parameter, \lstinline{Q}, which must match the signature \lstinline{PATTERN} (again, we use \lstinline{Q} to emphasize that the library provider does not make any assumptions about identifiers in scope; the client must pass in a particular module, e.g.
-\lstinline{P} above).
-The type annotation \lstinline{at Q.t} specifies that all expansions that arise from the application of this TSM to a module and a delimited form must necessarily be of type \lstinline{Q.t}, where \lstinline{Q} refers to that module.
+This \emph{TSM definition} first identifies the TSM as \lstinline{rx}, then specifies a module parameter, \lstinline{Q}, which quantifies over modules that match the signature \lstinline{RX} (again, we use \lstinline{Q} to emphasize that the library provider does not make any assumptions about identifiers in scope; the client must pass in a particular module, e.g.
+\lstinline{R} above).
+The \emph{type annotation} \lstinline{at Q.t} specifies that all expansions that arise from the application of this TSM to a module, \lstinline{Q}, and a delimited form must necessarily be of type \lstinline{Q.t}.
 
@@
 the parameter \lstinline{P} and the delimited form \lstinline{/SURLA|T|G|CEURL/} in the example above, the expansion is computed by calling the \emph{parse function} defined within braces.
-This parse function is a static function of sort \lstinline{Body -> Exp}.
-Both of these sorts are defined in the Verse \emph{prelude}, which is a set of definitions available ambiently.
-The sort \lstinline{Body} gives the static function access to the {body} of the delimited form (e.g.
-the characters in blue above) and the sort \lstinline{Exp}, mentioned also in the discussion of macros above,  encodes the abstract syntax of external expressions (there are also encodings of other syntactic classes, e.g.
-types, that can appear in expressions).
-The parse function must treat the TSM parameters parametrically, i.e.
+The parse function is a static function of type \lstinline{Body -> Exp}.
+Both of these types are defined in the Verse \emph{prelude}, which is a set of definitions available ambiently.
+The type \lstinline{Body} gives the static function access to the {body} of the delimited form (e.g.
+the characters in blue above) and the type \lstinline{Exp}, mentioned also in the discussion of macros above,  encodes the abstract syntax of expressions (there are also encodings of other syntactic classes, e.g.
+types and variables, that can appear in expressions).
+%The parse function must treat the TSM parameters parametrically, i.e.
 it does not have access to any values in the supplied module parameter.
@@
 %syntax pattern_bad[Q : PATTERN] at Q.t {
-%  static fn (body :: Body) :: Exp => 
+%  static fn (body : Body) : Exp => 
 %    if Q.flag then (* ... *) else (* ... *)
@@
 To support splicing syntax as described in Sec.
-\ref{sec:syntax-extension}, the parse function must be able to extract external subexpressions directly from the supplied body.
+\ref{sec:syntax-extension}, the parse function must be able to extract subexpressions directly from the supplied body.
 For example, consider the client code below:
@@
 (* TSMs can be partially applied and abbreviated *)
-let syntax pat = pattern P
-let val ssn = pat /SURL\d\d\d-\d\d-\d\d\d\dEURL/
-fun example_tsm(name: string) => 
-  pat /SURL@EURLnameSURL: %EURLssn/
+let syntax rx = rx R
+let val ssn = rx /SURL\d\d\d-\d\d-\d\d\d\dEURL/
+fun example_using_tsm(name: string) => 
+  rx /SURL@EURLnameSURL: %EURLssn/
 \end{lstlisting}
 The subexpressions \lstinline{name} and \lstinline{ssn} on the last line appear directly in the body of the delimited form, so we call them \emph{spliced subexpressions}.
-When the parse function determines that a subsequence of the body should be treated as a spliced subexpression, it can mark it as such and insert it directly into the expansion.
-For example, the expansion generated for the body of \lstinline{example_tsm} above would, if written concretely with spliced expressions marked, be:
-\begin{lstlisting}[numbers=none]
-Q.Seq(Q.Str(<spliced>(name)), Q.Seq(Q.Str ": ", <spliced>(ssn)))
-\end{lstlisting}
-The hygiene mechanism then ensures that only these marked portions of the generated expansion can refer to the variables at the use site, preventing inadvertent variable capture by the expansion.
-%For example, had the  would not be a valid expansion, because the  that are not inside spliced subexpressions:
+When the parse function determines that a subsequence of the body should be treated as a spliced subexpression (here, by recognizing the characters \lstinline{@} and \lstinline{%} followed by an  identifier),
+it can mark this subsequence as such.
+Such marked subsequence can appear as subexpressions directly within the expansion being generated (i.e.
+marked subsequences of the body are of type \lstinline{Exp}).
+For example, the expansion generated for the body of \lstinline{example_using_tsm} would, if written concretely with spliced expressions marked, be:
+\begin{lstlisting}[numbers=none]
+Q.Seq(Q.Str(spliced<name>), Q.Seq(Q.Str ": ", spliced<ssn>))
+\end{lstlisting}
+Hygiene is achieved by checking that only these marked portions of the generated expansion refer to the variables at the use site, preventing inadvertent variable capture by the expansion.
+In other words, the TSM cannot make any assumptions about variable names at the use site.%For example, had the  would not be a valid expansion, because the  that are not inside spliced subexpressions:
 %\begin{lstlisting}[numbers=none]
@@
 After checking that the expansion is valid, i.e.
-that it is hygienic and that the expansion is of the type specified by the TSM, the semantics removes the splicing markers just mentioned and substitutes  the actual module parameter \lstinline{P} for \lstinline{Q}, producing the final expansion, as expected:
-\begin{lstlisting}[numbers=none]
-P.Seq(P.Str(name), P.Seq(P.Str ": ", ssn))\end{lstlisting}
+that it is hygienic as just described and that the expansion is of the type specified by the TSM, the semantics removes the splicing markers just mentioned and removes the abstraction barrier by substituting  the actual module parameter \lstinline{R} for \lstinline{Q}, producing the final expansion, as expected:
+\begin{lstlisting}[numbers=none]
+R.Seq(R.Str(name), R.Seq(R.Str ": ", ssn))\end{lstlisting}
 %Put another way, the  elaboration logic must be valid in any context.
 
-\paragraph{Formalization} To introduce the formal specification of TSMs, let us consider the rule in the EL's semantics for handling TSM application to a delimited form.
-To simplify matters, we will first consider only unparameterized TSMs (i.e.
+\subsection{Formalization}\label{sec:tsm-formalism}
+\begin{figure}$\begin{array}{llllll}
+\textbf{variables} & \textbf{type variables} & \textbf{TSM variables} & \textbf{bodies}\\
+x & t & m & b\\~\end{array}$\\
+$\begin{array}{l}
+\textbf{types}\\
+\tau ::= t ~\vert~ \tau \rightharpoonup \tau ~\vert~ \forall t.\tau ~\vert~ \mu t.\tau ~\vert~ \mathsf{1}\\
+~\\
+\textbf{marked types}\\
+\dot{\tau} ::= t ~\vert~ \dot{\tau} \rightharpoonup \dot{\tau} ~\vert~ \forall t.\dot{\tau} ~\vert~ \mu t.\dot{\tau} ~\vert~ \mathsf{1} ~\vert~ \mathsf{spliced}(\tau)\\
+~\\
+\textbf{TSM expressions}\\
+\eta ::= m ~\vert~ \mathsf{syntax} \,@\, \tau\,\{\hat{e}\}
+\\~\\
+\textbf{unexpanded expressions}\\
+\hat{e} ::= {x} ~\vert~ \lambda x{:}\tau.\hat{e} ~\vert~ \hat{e}(\hat{e}) ~\vert~ \Lambda t.\hat{e} ~\vert~ \hat{e}\texttt{[}\tau\texttt{]} ~\vert~ \mathsf{fold}[t.\tau](\hat{e}) ~\vert~ \mathsf{unfold}(\hat{e}) ~\vert~ () ~\vert~ \mathsf{let~syntax}~m = \eta ~\mathsf{in}~\hat{e} ~\vert~ \eta\,\texttt{/}b\texttt{/}
+\\~\\
+\textbf{marked expressions}\\
+\dot{e} ::= x ~\vert~ \lambda x{:}{\dot{\tau}}.\dot{e} ~\vert~ \dot{e}(\dot{e}) ~\vert~ \Lambda t.\dot{e} ~|~ \dot{e}\texttt{[}\dot{\tau}\texttt{]} ~\vert~ \mathsf{fold}[t.\dot\tau](\dot{e}) ~\vert~ \mathsf{unfold}(\dot{e}) ~\vert~ () ~\vert~ \mathsf{spliced}(\hat{e})
+\\~\\
+\textbf{expanded expressions}\\
+e ::= x ~\vert~ \lambda x{:}{\tau}.e ~\vert~ e(e) ~\vert~ \Lambda t.e ~\vert~ e\texttt{[}\tau\texttt{]} ~\vert~ \mathsf{fold}[t.\tau](e) ~\vert~ \mathsf{unfold}(e) ~\vert~ ()\end{array}\\\vspace{8px}$
+~\\
+$\begin{array}{ll}
+\textbf{type formation contexts} & \textbf{typing contexts}\\
+\Delta ::= \emptyset ~\vert~ \Delta, t & \Gamma ::= \emptyset ~\vert~ \Gamma, x : \tau
+%~\\
+%\textbf{typing contexts}\\
+%\Gamma ::= \emptyset ~\vert~ \Gamma, x : \tau
+\end{array}$
+% $\begin{array}{rlcll}
+% \textbf{variable} & x \\
+% \textbf{delimited form body} & b \\
+% ~\\
+% \textbf{type}  & \tau & ::= & \tau \rightarrow \tau & \text{function type}\\
+% & & | & \mathsf{1} & \text{unit type}\\
+% ~\\
+% \textbf{TSM expression} & \psi & ::= & \mathsf{syntax} @ \tau\,\{ \sigma \} & \text{TSM definition}\\
+% ~\\
+% \textbf{external expression} & e & ::= & x & \text{variable}\\
+% & & | & \lambda x{:}\tau. e & \text{function}\\
+% & & | & e(e) & \text{function application}\\
+% & & | & () & \text{unit value}\\
+% & & | & \psi \texttt{/}b\texttt{/} & \text{TSM application}\\
+% ~\\
+% \textbf{translational type}\\
+% ~\\
+% \textbf{translational expression}\\
+% ~\\
+% \end{array}$
+\caption{Syntax of $\mathcal{L}^\text{TSM}$}
+\label{fig:lambda-tsm-syntax}
+\end{figure}
+To give a formal account of TSMs, we will now introduce a typed lambda calculus, $\mathcal{L}^\text{TSM}$.
+To simplify matters, this calculus supports only unparameterized TSMs (i.e.
 TSMs defined at a single type, e.g.
-the  datatype \lstinline{Pattern}, rather than a parameterized family of types):
+the datatype \lstinline{Pattern}, rather than a parameterized family of types).
+In the dissertation, we will introduce a second calculus, $\mathcal{L}^\text{PTSM}$, that extends $\mathcal{L}^\text{TSM}$ with support for parameterized families of types as described above (see Sec.
+\ref{sec:syntax-timeline}).
+The syntax of $\mathcal{L}^\text{TSM}$ is shown in Figure \ref{fig:lambda-tsm-syntax}.
+
+\subsubsection{Types and Expanded Expressions}
+\emph{Types}, $\tau$, and \emph{expanded expressions}, $e$, form a standard typed lambda calculus supporting partial functions, quantification over types, recursive types and for simplicity, a single base type, $\mathsf{1}$.
+The reader can consult any standard text covering typed programming languages for the necessary background (e.g.
+\emph{TAPL} \cite{tapl} or \emph{PFPL} \cite{pfpl}).
+We will reproduce a more detailed account of the semantics of the language in the dissertation, but for our present purposes, it suffices to recall the relevant judgement forms.
+The static  semantics of expanded expressions can be specified by judgements of the following form:
+
+$\begin{array}{ll}
+\textbf{Judgement Form} & \textbf{Description}\\
+\Delta \vdash \tau~\mathtt{type} & \text{$\tau$ is a valid type under type formation context $\Delta$}\\
+\Delta \vdash \Gamma~\mathtt{ctx} & \text{$\Gamma$ is a valid typing context under $\Delta$}\\
+\Delta~\Gamma \vdash e : \tau & \text{$e$ has type $\tau$ under $\Delta$ and $\Gamma$}
+\end{array}$
+
+\noindent
+The dynamic semantics can be specified by judgements of the following form:
+
+$\begin{array}{ll}
+\textbf{Judgement Form} & \textbf{Description}\\
+e \mapsto e' & \text{$e$ transitions to $e'$}\\
+e~\mathtt{val} & \text{$e$ is a value}
+\end{array}$
+\\\noindent
+We will write $e \mapsto^* e'$ for the reflexive, transitive closure of the transition relation and $e \Downarrow e'$ iff $e \mapsto^* e'$ and $e'~\mathtt{val}$.
+
+\subsubsection{Macro Expansion}
+Programs ultimately evaluate as expanded expressions, but programmers write programs as \emph{unexpanded expressions}, $\hat{e}$.
+Unexpanded expressions are typechecked and expanded simultaneously according to the rules defining the \emph{typed expansion judgement}:
+
+$\begin{array}{ll}
+\textbf{Judgement Form} & \textbf{Description}\\
+\Delta~\Gamma \vdash \hat{e} \leadsto e : \tau & \text{$\hat{e}$ expands to $e$ at type $\tau$ under $\Delta$ and $\Gamma$}
+\end{array}$
+
+Every form in the syntax of $e$ has a corresponding form in the syntax of $\hat{e}$ (cf.
+Figure \ref{fig:lambda-tsm-syntax}).
+For each rule in the static semantics of $e$, there is a corresponding typed expansion rule where the unexpanded and expanded forms correspond.
+For example, the rules for variables, functions and function application are shown below (the remaining such rules are analagous, but we omit them here for concision):
 \begin{mathpar}
-\inferrule[syn-aptsm]{
-    \vdash \psi ~@~ \sigma_\text{ty}~\{\sigma_\text{parser}\}\\
-    b \downarrow \sigma_\text{body}\\
-    \sigma_\text{parser}(\sigma_\text{body}) \Downarrow \sigma_\text{exp}\\
-    \sigma_\text{exp} \uparrow e_\text{exp}\\\\
-    \Gamma; \emptyset \vdash e_\text{exp} \Leftarrow \sigma_\text{ty} \leadsto \iota_\text{trans}
+\inferrule[T-U-var]{ }{\Delta~\Gamma, x : \tau \vdash x \leadsto x : \tau}
+
+\inferrule[T-U-abs]{
+	\Delta \vdash \tau_1~\mathtt{type}\\
+	\Delta~\Gamma, x:\tau \vdash \hat{e} \leadsto e : \tau'
 }{
-    \Gamma_\text{out}; \Gamma \vdash \mathbf{aptsm}(\psi; b) \Rightarrow \sigma_\text{ty} \leadsto \iota_\text{trans}
+	\Delta~\Gamma \vdash \lambda x{:}\tau.\hat{e} \leadsto \lambda x{:}\tau.e : \tau \rightharpoonup \tau'hh
+}
+
+\inferrule[T-U-ap]{
+	\Delta~\Gamma \vdash \hat{e}_1 \leadsto e_1 : \tau \rightharpoonup \tau' \\
+	\Delta~\Gamma \vdash \hat{e}_2 \leadsto e_2 : \tau
+}{
+	\Delta~\Gamma \vdash \hat{e}_1(\hat{e}_2) \leadsto e_1(e_2) : \tau'
 }
 \end{mathpar}
-Looking at the conclusion of the rule, we see that TSM application takes the abstract form $\mathbf{aptsm}(\psi; b)$.
-The metavariable $\psi$ ranges over \emph{TSM expressions} and $b$ ranges over {bodies} of delimited forms (i.e.
-the sequences of characters between the delimiters in the examples above).
-TSM expressions consist only of TSM definitions here (cf.
-the definition of \lstinline{pattern} above), though when we formalize parameterized TSMs, we will distinguish TSM definitions from TSM expressions, which involve parameter application.%Note that for simplicity in the formal system, we have tacitly assumed that TSM names are drawn from a set that is disjoint from the set of variable names, so that function application and TSM application desugar to the distinct abstract forms.
-%In practice, we would use a common set of names and include a disambiguation step, but making the assumption of disjoint names allows us to avoid specifying this.
-%In practice, there would be an additional step that determined whether a function or a TSM was being applied based on the current context,
-
-Note also that we omit various contexts relevant only to other Verse primitives.
-We include only \emph{typing contexts}, $\Gamma$, which map variables to types in the standard way.
-We describe why there are two contexts -- the \emph{outer context} and the \emph{current context} -- below.
-
+There are two forms in the syntax of $\hat{e}$ that have no corresponding form in the syntax of $e$.
+The first allows the programmer to bind a \emph{TSM variable}, $m$, to a \emph{TSM expression}, $\eta$.
+TSM expressions are either TSM variables or \emph{TSM definitions}.
+Substitution for TSM variables is performed statically, so we only need a rule for the case where the TSM variable is being bound to a TSM definition:
+\begin{mathpar}
+\inferrule[T-U-TSM-let]{
+	\Delta \vdash \mathsf{syntax}~@~\tau~\{\hat{e}_\text{parse}\}~\mathtt{tsm}\\
+	\Delta~\Gamma \vdash [\mathsf{syntax}~@~\tau~\{\hat{e}_\text{parse}\}/m]\hat{e} \leadsto e : \tau'
+}{
+	\Delta~\Gamma \vdash \mathsf{let}~\mathsf{syntax}~m=\mathsf{syntax}~@~\tau~\{\hat{e}_\text{parse}\}~\mathsf{in}~\hat{e} \leadsto e : \tau'
+}
+\end{mathpar}
+The first premise checks that the TSM definition is valid. It is defined by the following rule:
+\begin{mathpar}
+\inferrule[TSM-OK]{
+	\Delta \vdash \tau~\mathtt{type}\\
+	\emptyset~\emptyset \vdash \hat{e}_\text{parse} \leadsto e_\text{parse} : \mathsf{Body} \rightharpoonup \mathsf{Exp}
+}{
+	\Delta \vdash \mathsf{syntax}~@~\tau~\{\hat{e}_\text{parse}\}~\mathtt{tsm}
+}
+\end{mathpar}
+The first premise of (TSM-OK) checks that the type is valid.
+The second premise typechecks and expands the parse function, which must be closed.
+We discuss the abbreviated types $\mathsf{Body}$ and $\mathsf{Exp}$ below.
+
+The second form in the syntax of $\hat{e}$ that has no corresponding form in the syntax of $e$ is the form for TSM application to a delimited body, $\eta\,\texttt{/}b\texttt{/}$.
+Again because substitution for TSM variables is performed statically, we only need a rule for the case where $\eta$ is a TSM definition:
+\begin{mathpar}
+\inferrule[T-U-TSM-ap]{
+	\Delta \vdash \tau~\mathtt{type}\\
+	\emptyset~\emptyset \vdash \hat{e}_\text{parse} \leadsto e_\text{parse} : \mathsf{Body} \rightharpoonup \mathsf{Exp}\\\\
+	b \downarrow e_\text{body} : \mathsf{Body}\\
+	e_\text{parse}(e_\text{body}) \Downarrow e_\text{exp}\\
+	e_\text{exp} : \mathsf{Exp} \uparrow \dot{e}_\text{exp}\\\\
+	\Delta~\Gamma; \emptyset~\emptyset \vdash \dot{e}_\text{exp} \leadsto e : \tau
+}{
+	\Delta~\Gamma \vdash \mathsf{syntax}~@~\tau~\{\hat{e}_\text{parse}\}~\texttt{/}b\texttt{/} \leadsto e : \tau
+}
+\end{mathpar}
 The premises can be understood as follows, in order:
 \begin{enumerate}
-\item The first premise can be pronounced ``TSM expression $\psi$ defines a valid TSM at type $\sigma_\text{ty}$ with parse function $\sigma_\text{parser}$''.
-The ``outputs'' of this judgement are the type $\sigma_\text{ty}$ (of sort $\mathsf{Ty}$) and the static parse function, $\sigma_\text{parser}$ (of sort $\mathsf{ParseStream} \rightarrow \mathsf{Exp}$), defined by the TSM.% We must omit the rules governing this judgement here (a complete specification of this judgement represents the main piece of remaining work, see below).
-% Application of a parameterized TSM to its parameters is handled by this judgement (not shown).
-\item The second premise creates an encoding of the body of the delimited form, $\sigma_\text{body}$, of sort $\mathsf{Body}$.
-\item The third premise applies the parse function to the parse stream to generate a static representation of the expansion, $\sigma_\text{exp}$ (of sort $\mathsf{Exp}$).% Note that this is ``metaprogramming'' -- a static expression in the external semantics to generate an encoding of another external term.
-\item The fourth premise decodes $\sigma_\text{exp}$, producing the expansion itself, $e_\text{exp}$.
-\item The final premise \emph{validates} the expansion by analyzing it against the type $\sigma_\text{ty}$ specified by the TSM.
-The {current} typing context is ``saved'' for use when a spliced subexpression is encountered during this process by setting it as the new \emph{outer context}, as indicated.
-
-Variables in the outer context are not directly available to the expansion, e.g. they are ignored by the (syn-var) rule:
+\item The first premise checks that the type specified by the TSM is valid.
+\item The second premise typechecks and expands the parse function.
+\item The third premise encodes the body, $b$, as a term, $e_\text{body}$ of type $\mathsf{Body}$ (we will give more details on how bodies are encoded in the dissertation, but omit the definition of $\mathsf{Body}$ here for concision).
+\item The fourth premise applies the expanded parse function to the encoding of the body to produce an encoding of the expansion, $e_\text{exp}$, of  type $\mathsf{Exp}$. 
+\item Values of type $\mathsf{Exp}$ map onto \emph{marked expressions}, $\dot{e}$.
+Per Figure \ref{fig:lambda-tsm-syntax}, marked expressions can contain variables, type variables and marked types, $\dot{\tau}$, so there is also a mapping from values of types abbreviated $\mathsf{Var}$, $\mathsf{TVar}$ and $\mathsf{Type}$ onto variables, type variables, and {marked types}, respectively.
+We also omit the full definitions of these types for concision (cf.
+the SML/NJ Visible Compiler library \cite{SML/VisibleCompiler} for an example of such an encoding of the abstract syntax of a language).
+These mappings are specified by the \emph{expansion decoding judgements}:
+
+$\begin{array}{ll}
+\textbf{Judgement Form} & \textbf{Description}\\
+e : \mathsf{Var} \uparrow x & \text{$e$ decodes to variable $x$.}\\
+e : \mathsf{TVar} \uparrow t & \text{$e$ decodes to type variable $t$.}\\
+e : \mathsf{Type} \uparrow \dot{\tau} & \text{$e$ decodes to marked type $\dot{\tau}$.}\\
+e : \mathsf{Exp} \uparrow \dot{e} & \text{$e$ decodes to marked expression $\dot{e}$.}\\
+\end{array}$
+
+The fifth premise decodes $e_\text{exp}$ to produce the \emph{marked expansion}, $\dot{e}_\text{exp}$. 
+\item The final premise valides the expansion by checking the marked expansion against the type specified by the TSM, and generates the final expansion, $e$, according to the rules defining the \emph{expansion validation judgements}:
+
+$\begin{array}{ll}
+\textbf{Judgement Form} & \textbf{Description}\\
+\Delta_\text{out}; \Delta \vdash \dot{\tau} \leadsto \tau~\mathtt{type} & \text{Marked type $\dot{\tau}$ expands to $\tau$ under outer context $\Delta_\text{out}$}\\
+& \text{and current context $\Delta$.}\\
+\Delta_\text{out}~\Gamma_\text{out}; \Delta~\Gamma \vdash \dot{e} \leadsto e : \tau & \text{Marked expression $\dot{e}$ expands to $e$ at type $\tau$ under outer}\\
+& \text{contexts $\Delta_\text{out}$ and $\Gamma_\text{out}$ and current contexts $\Delta$ and $\Gamma$.}
+\end{array}$
+
+Each form in the syntax of expanded expressions has a corresponding form in the syntax of marked expressions (cf.
+Figure \ref{fig:lambda-tsm-syntax}).
+For each rule in the static semantics of $e$, there is a corresponding expansion validation rule where the marked and expanded forms correspond.
+Only the current contexts are examined or extended by these rules.
+For example, the expansion validation rules for variables, functions and function application are shown below (the remaining such rules are analagous, but we omit them for concision):
 \begin{mathpar}
-\inferrule[syn-var]{ }{
-    \Gamma_\text{out}; \Gamma, x : \sigma \vdash x \Rightarrow \sigma \leadsto x
+\inferrule[T-M-var]{ }{\Delta_\text{out}~\Gamma_\text{out}; \Delta~\Gamma, x : \tau \vdash x \leadsto x : \tau}
+
+\inferrule[T-M-abs]{
+	\Delta_\text{out}; \Delta \vdash \dot{\tau} \leadsto \tau ~\mathtt{type}\\
+	\Delta_\text{out}~\Gamma_\text{out}; \Delta~\Gamma, x : \tau \vdash \dot{e} \leadsto e : \tau'
+}{
+	\Delta_\text{out}~\Gamma_\text{out}; \Delta~\Gamma \vdash \lambda x{:}\dot{\tau}.\dot{e} \leadsto \lambda x{:}\tau.e : \tau \rightharpoonup \tau'
+}
+
+\inferrule[T-M-app]{
+	\Delta_\text{out}~\Gamma_\text{out}; \Delta~\Gamma \vdash \dot{e}_1 \leadsto e_1 : \tau \rightharpoonup \tau'\\
+	\Delta_\text{out}~\Gamma_\text{out}; \Delta~\Gamma \vdash \dot{e}_2 \leadsto e_2 : \tau
+}{
+	\Delta_\text{out}~\Gamma_\text{out}; \Delta~\Gamma \vdash \dot{e}_1(\dot{e}_2) \leadsto e_1(e_2) : \tau'
 }
@@
 
-Outer contexts are switched back in when encountering marked spliced expressions:
+The purpose of the outer contexts is to ``remember'' the context that the macro application appeared in so that spliced subexpressions extracted from the body, which are marked with the form $\mathsf{spliced}(\hat{e})$, can be checked appropriately:
 \begin{mathpar}
-\inferrule[syn-spliced]{
-    \emptyset; \Gamma_\text{out} \vdash e \Rightarrow \sigma \leadsto \iota
+\inferrule[T-M-spliced]{
+	\Delta_\text{out}~\Gamma_\text{out} \vdash \hat{e} \leadsto e : \tau
 }{
-    \Gamma_\text{out}; \Gamma \vdash \mathbf{spliced}(e) \Rightarrow \sigma \leadsto \iota
-}
-
-\inferrule[ana-spliced]{
-    \emptyset; \Gamma_\text{out} \vdash e \Leftarrow \sigma \leadsto \iota
-}{
-    \Gamma_\text{out}; \Gamma \vdash \mathbf{spliced}(e) \Leftarrow \sigma \leadsto \iota
+	\Delta_\text{out}~\Gamma_\text{out}; \Delta~\Gamma \vdash \mathsf{spliced}(\hat{e}) \leadsto e : \tau
 }
@@
 
-This premise of (syn-aptsm) thus enforces both type safety and hygiene.
-
+The current contexts are initially empty when checking the marked expansion generated by the parse function, so we achieve hygiene: the expansion cannot make any assumptions about the variables available in the outer context.
 %The translation of the elaboration becomes the translation in the conclusion of the rule.
@@
 %We plan to give a more complete account of all of these judgements in the dissertation.
-\subsubsection{Type-Specific Languages (TSLs)}\label{sec:tsls}
-With TSMs, we achieved the conventional syntax for regular expression patterns within delimiters, but we still explicitly had to apply the TSM and provide the module parameter at each application site.
-To further lower the syntactic cost of using TSMs, so that it compares to the syntactic cost of derived syntax built in primitively, Verse also allows library providers to associate a TSM directly with an abstract or declared type (or type constructor, more generally).
-When the semantics encounters a delimited form not prefixed by a TSM name, the TSM associated with the type it is being analyzed against (i.e.
-that type's \emph{type-specific language}) is implicitly applied
-
-For example, a module \lstinline{P} can associate \lstinline{pattern} with \lstinline{P.t} by qualifying the definition of the opaquely ascribed module it is defined by as follows:
-\begin{lstlisting}[numbers=none]
-module P = mod {
+\subsection{Limitations}\label{sec:tsms-limitations}
+%\todo{revise this section}The main limitation of this approach is that it 
+Though expansion validation ensures that an incorrect parser cannot threaten type safety,  proving type safety inductively is subtle because the typed expansion judgement and the expansion validation judgement are mutually defined.
+We must come up with a decreasing metric to ensure that induction is well-founded.
+We will formally define such a metric in the dissertation, but informally, notice that we only induct on spliced terms derived from the body, which is itself a subterm of the original unexpanded expression.
+
+To provide the strongest metatheoretic guarantees, we must have that the static subset of the language is total.
+Allowing non-termination in the parse function (as we have done in the formalism above, by not distinguishing a static subset at all) would cause typed expansion to become undecidable (typechecking expanded expressions would remain decidable, however).
+Related concerns would arise if the static subset supported external (e.g.
+I/O) effects -- one would generally not want the mere act of running the typechecker on a program to have an arbitrary effect on, for example, the programmer's file system.
+Mutation effects could also be problematic if we allow static values to be ``promoted'' into expansions (a feature we did not specify above, but will return to in the dissertation).
+
+Another limitation is that TSMs as we have described them only capture idioms that occur within a single parameterized family of types,  not idioms that might span many (or all) types, e.g.
+control flow idioms.
+In fact, this is intentional -- neither humans nor tools need to examine the expansion to determine what type it must necessarily have.
+We will, however, discuss other points in the design space in the dissertation (based on some of the work in a recent paper we have published \cite{sac15}).
+ %, the essential ideas are mostly simply conveyed by sticking to formal specifications only at the level of expressions.
+
+\subsection{Typed Pattern Syntax Macros (TPSMs)}\label{sec:pattern-tsms}
+TSMs as we have described them so far decrease the syntactic cost of introducing a value at a specified type.
+In full-scale functional languages like ML, one typically deconstructs a value using \emph{nested pattern matching}.
+For example, let us return to the definition of the datatype \lstinline{Rx} shown at the beginning of Sec.
+\ref{sec:examples}.
+We can pattern match over a value, \lstinline{r}, of type \lstinline{Rx} using Verse's \lstinline{match} construct like this:
+\begin{lstlisting}[numbers=none]
+match r with 
+    Seq(Str(name), Seq(Str ": ", ssn)) => display name ssn
+  |  _ => raise Invalid
+\end{lstlisting}
+In a functional language with primitive support for regular expression pattern syntax, we would expect to be able to write this example more concisely using the splicing forms discussed in Sec.
+\ref{sec:tsms-by-example}:
+\begin{lstlisting}[numbers=none]
+match r with 
+    /SURL@EURLnameSURL: %EURLssn/ => display name ssn
+  | _ => raise Invalid
+\end{lstlisting}
+
+Patterns are not expressions, so we cannot simply use a TSM defined at type \lstinline{Rx} in a pattern.
+To address this, we must extend our language with support for typed pattern syntax macros (TPSMs).
+TPSMs are entirely analagous to TSMs, differing primarily in that the expansions they generate are patterns, rather than expressions.
+Assuming the abstract syntax of patterns is encoded by the type \lstinline{Pat} (analagous to \lstinline{Exp}), we can define a TPSM at type \lstinline{Rx} as follows:
+\begin{lstlisting}[numbers=none]
+pattern syntax rx at Rx {
+	static fn (body : Body) : Pat => 
+	  (* regex pattern parser here *)
+}
+\end{lstlisting}
+
+Using this TPSM, we can rewrite our example as follows:
+\begin{lstlisting}[numbers=none]
+match r with 
+    rx /SURL@EURLnameSURL: %EURLssn/ => display name ssn
+  | _ => raise Invalid
+\end{lstlisting}
+To ensure that the client of the TPSM need not ``guess at'' what variables are bound by the pattern, variables (e.g.
+\lstinline{name} and \lstinline{ssn} here) can only appear in spliced subpatterns (just as variables bound at the use site can only appear in spliced subexpressions when using TSMs).
+We leave a formal account of TPSMs (in a reduced calculus that features simple pattern matching) as work that remains to be completed (see Sec.
+\ref{sec:syntax-timeline}).
+
+ML does not presently support pattern matching over values of an abstract data type.
+However, there have been proposals for adding support for pattern matching over abstract data types defined by modules having a ``datatype-like'' shape, e.g.
+those that define a case analysis function like the one specified by \lstinline{RX}, shown in Sec.
+\ref{sec:examples}.
+We leave further discussion of such a facility and of parameterized TPSMs also as remaining work (see Sec.
+\ref{sec:syntax-timeline}).
+
+\section{Type-Specific Languages (TSLs)}\label{sec:tsls}
+With TSMs, library providers can control the expansion of arbitrary syntax that appears between delimiters, but clients must explicitly identify the TSM and provide the required type and module parameters at each use site.
+To further lower the syntactic cost of using TSMs, so that it compares to the syntactic cost of derived syntax built in primitively (e.g.
+list syntax), we will now discuss how Verse allows library providers to define \emph{type-specific languages} (TSLs) by associating a TSM directly with an abstract type or datatype.
+When the type system encounters a delimited form not prefixed by a TSM name, it applies the TSM associated with the type it is being analyzed against implicitly.
+
+For example, a module \lstinline{P} can associate the TSM \lstinline{rx} defined in the previous section with the abstract type \lstinline{R.t} by qualifying the definition of the sealed module it is defined by as follows:
+\begin{lstlisting}[numbers=none]
+module R = mod {
   type t = (* ... *)
   (* ... *)
-} :> PATTERN with syntax pattern
-\end{lstlisting}
-This definition associates the TSM \lstinline{pattern P} with the abstract type \lstinline{P.t}.
-As such, the following function is equivalent to \lstinline{example_tsm} (note the return type annotation, which is now necessary):
-\begin{lstlisting}[numbers=none]
-fun example_tsl(name : string) : P.t => 
-  /SURL@EURLnameSURL: %EURLssn/
-\end{lstlisting}
-
-As another example, we can use TSLs to express derived list syntax. For example, if we use a datatype, we can define the TSL directly upon declaring the list type:
+} :> RX with syntax rx
+\end{lstlisting}
+More generally, when sealing a module expression against a signature, the programmer can specify, for each abstract type that is generated, at most one previously defined TSMs.
+This TSM must take as its first parameter the module being sealed.
+
+The following function has the same expansion as \lstinline{example_using_tsm} but, by using the TSL just defined, it is more concise.
+Notice the return type annotation, which is necessary to ensure that the TSL can be unambiguously determined:
+\begin{lstlisting}[numbers=none]
+fun example_using_tsl(name : string) : R.t => /SURL@EURLnameSURL: %EURLssn/
+\end{lstlisting}
+
+As another example, let us consider the standard list datatype. We can use TSLs to express derived list syntax, for both expressions and patterns:
 \begin{lstlisting}[numbers=none]
 datatype list('a) { Nil | Cons of 'a * list('a) } with syntax {
-  static fn (body :: ParseStream) => 
+  static fn (body : Body) => 
     (* ... comma-delimited spliced exps ... *)
+} with pattern syntax {
+  static fn (body : Body) : Pat => 
+    (* ... list pattern parser ... *)
 }
 \end{lstlisting}
-Together with the TSL for patterns, this allows us to write a list of patterns like this:
-\begin{lstlisting}[numbers=none]
-let val x : list(P.t) = [/SURL\dEURL/SHTML, EHTML/SURL\d\dEURL/SHTML, EHTML/SURL\d\d\dEURL/]
+Together with the TSL for regular expression patterns, this allows us to write lists like this:
+\begin{lstlisting}[numbers=none]
+let val x : list(R.t) = [/SURL\dEURL/SHTML, EHTML/SURL\d\dEURL/SHTML, EHTML/SURL\d\d\dEURL/]
 \end{lstlisting}
@@
 
-\paragraph{Abstract Types}
-Let us examine how this mechanism works for abstract types in more detail.
-When sealing a module expression with a signature, the programmer can specify a previously defined TSMs for each abstract type that arises from the ascription.
-Each must be parameterized by a module satisfying the specified signature.
-
-For parameterized modules, the situation is similar. For example, consider a trivially parameterized module that creates modules sealed against \lstinline{PATTERN}:
+\subsection{Parameterized Modules}
+TSLs can be associated with abstract types that are generated by parameterized modules (i.e.
+generative functors in Standard ML) as well.
+For example, consider a trivially parameterized module that creates modules sealed against \lstinline{RX}:
 \begin{lstlisting}[numbers=none]
@@
   (* ... *)
-} :> PATTERN with syntax pattern 
-\end{lstlisting}
-Each application of \lstinline{F} generates a distinct abstract type. The semantics associates the appropriately parameterized TSM with each of these:
-\begin{lstlisting}[numbers=none]
-module F1 = F() (* F1.t has TSL pattern(F1) *)
-module F2 = F() (* F2.t has TSL pattern(F2) *)
-\end{lstlisting}
-
-For a parameterized module that takes non-trivial parameters, we need only define the TSM to also take the same parameters and apply them explicitly within the definition of the parameterized module.
-For example, let us define a parameterized module \lstinline{G : A -> B}:
+} :> RX with syntax rx 
+\end{lstlisting}
+Each application of \lstinline{F} generates a distinct abstract type. The semantics associates the appropriately parameterized TSM with each of these as they are generated:
+\begin{lstlisting}[numbers=none]
+module F1 = F() (* F1.t has TSL rx(F1) *)
+module F2 = F() (* F2.t has TSL rx(F2) *)
+\end{lstlisting}
+
+As a more complex example, let us define two signatures, \lstinline{A} and \lstinline{B}, a TSM \texttt{\$G} and a parameterized module \lstinline{G : A -> B}:
 \begin{lstlisting}[numbers=none,mathescape=|]
@@
 \end{lstlisting}
-Note that within the definition of \lstinline{G}, type \lstinline{u} does not have a TSL available to it.
-
-TSL lookup respects type equality, so any synonyms of a type with a TSL will also have that TSL.
+Both \lstinline{G} and \texttt{\$G} take a parameter \lstinline{M : A}.
+We associate the partially applied TSM \texttt{\$G(M)} with the abstract type that \lstinline{G} generates.
+Again, this satisfies the requirement that one must be able to apply the TSM being associated with the abstract type to the module being sealed.
+
+Only fully abstract types can have TSLs associated with them.
+Within the definition of \lstinline{G}, type \lstinline{u} does not have a TSL available to it because it is synonymous to \lstinline{M.t}.
+More generally, TSL lookup respects type equality, so any synonyms of a type with a TSL will also have that TSL.
 We can see this in the following example, where the type \lstinline{u} has a different TSL associated with it inside and outside the definition of the module \lstinline{N}:
@@
             so N.u has no TSL in the outer scope *)
-\end{lstlisting}
-
-\subsubsection{Limitations}\label{sec:tsms-limitations}
-%\todo{revise this section}The main limitation of this approach is that it 
-Though expansion validation ensures that an incorrect parser cannot threaten type safety,  proving type safety inductively is subtle because the expansion is not a subexpression of the expression it will replace.
-There is a ``simple'' solution to this problem: we impose the restriction that the expansion cannot itself use TSMs or TSLs.
-This would be somewhat awkward in practice, so we plan to briefly discuss other more flexible solutions as well.
-
-To provide the strongest metatheoretic guarantees, we must have that the SL is a total functional language.
-Allowing non-termination in the SL would cause typechecking of the EL to become undecidable because macro expansion occurs during typechecking (typechecking the IL would remain decidable, however).
-Related concerns would arise if the static language supported external (e.g.
-I/O) effects -- one would generally not want the mere act of parsing and typechecking a program to have an arbitrary effect on, for example, the programmer's file system.
-Mutation effects could, however, be introduced, because the static language is distinct from the external language.
-No static values can be ``promoted'' into the external language.
-
-Another limitation is that TSMs as we have described them only capture idioms that occur within a single parameterized family of types,  not idioms that might span many (or all) types, e.g.
-control flow idioms.
-In fact, this is intentional -- neither humans nor tools need to examine the expansion to determine what type it must necessarily have.
-We will, however, discuss other points in the design space in the dissertation.
-
-We have discussed only expression syntax here, but have not considered pattern matching.
-In fact, we conjecture that it is completely straightforward to extend this work to allow a TSM to define both an expression parser and a pattern parser.
-Similarly, one could imagine defining TSMs at the level of modules, and at the level of types.
-We plan to discuss these in the dissertation.
-Formally specifying these will be left to future work, as it would introduce a rather large amount of technical overhead for minimal conceptual gain.
-%, the essential ideas are mostly simply conveyed by sticking to formal specifications only at the level of expressions.
-
-\subsection{Timeline and Milestones}\label{sec:syntax-timeline}
-We described and gave a more detailed formal specification of TSMs in a recently published paper \cite{sac15}, and TSLs in a paper published last year \cite{TSLs}, both in the context of the Wyvern language.
+val z : N.u = /asdf/ (* ERROR: no TSL for type N.u *)
+\end{lstlisting}
+
+\subsection{Formalism}
+A formal specification of TSLs in a language that supports only non-parametric datatypes is available in a paper published in ECOOP 2014 \cite{TSLs}.
+We will add support for parameterized TSLs in the dissertation (see Sec.
+\ref{sec:syntax-timeline}).
+
+\section{Timeline and Milestones}\label{sec:syntax-timeline}
+We described and gave a more detailed formal specification of TSMs in a recently published paper \cite{sac15}, and TSLs in a paper published last year \cite{TSLs}, both in the context of the Wyvern programming language.
 The mechanics of defining a TSM, statically invoking the parse function, expansion encoding/decoding and hygienic expansion validation have all been detailed there.
-In the dissertation, I plan to first present the formal system closely following this presentation, updating it only to ensure that it is consistent with the description above and with the work in the next section.
-The rule (syn-aptsm) above represents the first and most substantial step of this plan.
-This is complete on paper, so I only need to typeset it.
-
-Wyvern does not specify an ML-style module system (type declarations are the only form of type generation) and at the time, it did not support parameterized types, so to actually support the examples we have given in this section, we then need to update the formalization to handle parameterized TSMs.
-I plan to flesh this out in the course of writing the dissertation, and I anticipate it requiring a further 1 month of my time after the proposal is complete.
-I propose the following concrete milestones:
-
-\begin{enumerate}
-\item equip the EL with an appropriately structured module context \textbf{(complete on paper, needs to be typeset)}
-\item formally specify the syntax of parameterized TSM expressions \textbf{(complete on paper, needs to be typeset)}
-\item formally specify an updated version of the TSM expression validation judgement that appeared as the first premise of (syn-aptsm) above \textbf{(complete on paper, needs to be typeset)}
-\item update the metatheory established in the work cited above to handle this
-\end{enumerate}
-
-To validate our claims of expressiveness, I plan to give several more examples drawn from existing languages.
+
+In the dissertation, I plan to first present the formal system $\mathcal{L}^\text{TSM}$.
+The account in Sec.
+\ref{sec:tsm-formalism} represents the first and most substantial step of this plan.
+This is complete on paper, so I only need to typeset the omitted details, including the metatheory.
+I plan to finish this by the \textbf{end of October}.
+This system will serve as the basis for several extensions that capture other constructs described above, so I will ask the committee to read this section in detail once it is complete.
+
+The first extension will add support for simple TSLs over unparameterized datatypes.
+This has already been written up in \cite{TSLs}, so I plan to have this in final form in the \textbf{first week of November}.
+
+The next extension will add support for pattern matching over these datatypes, and TPSMs.
+I will base this extension on the account of pattern matching given in \emph{PFPL} \cite{pfpl}.
+I plan to have a rough sketch of how this should work by \textbf{late-November}, in consultation with the committee, and a final write-up of it by \textbf{mid-December}.
+
+The final extension will add support for type and module parameters.
+To do so, I will assume that the available module paths and their signatures are being tracked by a module context in the static semantics, without detailing how the module language populates this context (i.e.
+I will only specify the ``core language'').
+I plan to sketch this out and discuss it with my committee also \textbf{during November} and finish writing the details up by the \textbf{last week of December}.
+
+Over the course of the \textbf{Spring semester}, I plan to write the remaining sections (e.g.
+additional background, related and future work sections, and additional examples), flesh out the proofs of minor lemmas and respond to the committee's requests for edits.
+Based on this timeline, I plan to defend before the \textbf{end of the Spring semester}.
+
+My current draft of the dissertation document will be available on GitHub:
+
+\url{https://github.com/cyrus-/papers/blob/master/thesis/omar-thesis.pdf}
+%To validate our claims of expressiveness, I plan to give several more examples drawn from existing languages.
 I will include those we have worked out in the papers just cited as well as additional examples that highlight the benefits of integration with Verse's ML-based module system.
-For example, I will discuss implementing Haskell's primitive ``do'' syntax using TSLs together with a modular encoding of monads (as described on Prof.
+%For example, I will discuss implementing Haskell's primitive ``do'' syntax using TSLs together with a modular encoding of monads (as described on Prof.
 Harper's blog \cite{SML/Monads}).
@@
 as in Scala \cite{shabalin2013quasiquotes} or Lisp \cite{Bawd99a}) can be seen as modes of use of TSMs/TSLs.
-I will also briefly summarize the corpus analysis we conducted in \cite{TSLs} where we found a large number of  situations in large publicly available Java libraries where dynamic string parsing was apparently being used to control syntactic cost.
-I anticipate that writing these examples will take an additional 1 week of work, bringing the total for this portion of the dissertation up to about 5 weeks of writing (plus time for review by the committee).
+%I will also briefly summarize the corpus analysis we conducted in \cite{TSLs} where we found a large number of  situations in large publicly available Java libraries where dynamic string parsing was apparently being used to control syntactic cost.
+I anticipate that writing these examples will take an additional 1 week of work, which I will do after finishing the technical work above.
+%bringing the total for this portion of the dissertation up to about 5 weeks of writing (plus time for review by the committee).
 %Adding in the time I will need to propose and some buffer time for unanticipated issues, my goal is to have this portion of the dissertation complete by the end of Summer 2015.
@@
 
-\section{Modularly Programmable Type Structure}\label{sec:metamodules}
-In the previous section, we assumed that the EL had an otherwise standard, ML-like type structure and considered only TSMs and TSLs as departures from ML.
+% \section{Modularly Programmable Type Structure}\label{sec:metamodules}
+% In the previous section, we assumed that the EL had an otherwise standard, ML-like type structure and considered only TSMs and TSLs as departures from ML.
 In fact, constructs like datatypes (shown briefly in use in the previous section), tuples, records and objects are not built primitively into Verse.
@@
 
-This section is organized like Sec.
+% This section is organized like Sec.
 \ref{sec:syntax}.
@@
 
-\subsection{Motivating Example: Labeled Tuples and Regular Strings}\label{sec:metamodules-example}\label{sec:metamodules-motivating-examples}
-As a simple introductory example showing the sorts of types and operators we might like to have available to us as clients of a language, let us define a simple \emph{labeled tuple type} for conference papers:
-\begin{lstlisting}[numbers=none]
-let type Paper = ltuple {SPCT
-   title : EPCTrstring /SURL.+EURL/
-    SPCTconf : EPCTrstring /SURL([A-Z]+) (\d\d\d\d)EURL/
-}
-\end{lstlisting}
-The \lstinline{let type} construction defines a synonym, \lstinline{Paper}, for a type constructed by applying the \emph{type constructor} (or \emph{tycon}) \lstinline{ltuple} to a \emph{component specification}, which is a  statically valued (cf.
+% \subsection{Motivating Example: Labeled Tuples and Regular Strings}\label{sec:metamodules-example}\label{sec:metamodules-motivating-examples}
+% As a simple introductory example showing the sorts of types and operators we might like to have available to us as clients of a language, let us define a simple \emph{labeled tuple type} for conference papers:
+% \begin{lstlisting}[numbers=none]
+% let type Paper = ltuple {SPCT
+%    title : EPCTrstring /SURL.+EURL/
+%     SPCTconf : EPCTrstring /SURL([A-Z]+) (\d\d\d\d)EURL/
+% }
+% \end{lstlisting}
+% The \lstinline{let type} construction defines a synonym, \lstinline{Paper}, for a type constructed by applying the \emph{type constructor} (or \emph{tycon}) \lstinline{ltuple} to a \emph{component specification}, which is a  statically valued (cf.
 Sec.
@@
 
-The first component is labeled \lstinline{title} and specifies the \emph{regular string type} \lstinline{rstring /SURL.+EURL/}.
+% The first component is labeled \lstinline{title} and specifies the \emph{regular string type} \lstinline{rstring /SURL.+EURL/}.
 Regular string types are constructed by applying the tycon \lstinline{rstring} to a statically valued regular expression pattern, again written here using conventional concrete syntax like that discussed in the previous section.
@@
 
-% Alternatively, we could have used the following construction to declare \lstinline{Paper} \emph{generatively}:
-% \begin{lstlisting}[numbers=none]
-% lprodtype Paper {
-%   SPCTtitle : EPCTrstring /SURL.+EURL/
-%    SPCTconf : EPCTrstring /SURL([A-Z]+) (\d\d\d\d)EURL/
-% }
-% \end{lstlisting}
-% The difference is in how type equality is handled. Two labeled product types with the same signature are equal, unless declared generatively, in which case type equality is nominal.
-
-Clients can introduce regular strings in analytic position using standard string literal syntax, as long as the body of the literal is in the corresponding regular language. 
-Clients can introduce a value of type \lstinline{Paper} in an analytic position in one of two ways. We can omit the labels and provide component values positionally:
-\begin{lstlisting}[numbers=none]
-let val exmpl : Paper = {"SUSAn Example PaperEUS"SCSS, ECSS"SUSEXMPL 2015EUS"}
-\end{lstlisting}
-Alternatively, we can include the component labels explicitly for readability or if we want to give the components in an alternative order:
-\begin{lstlisting}[numbers=none]
-let val exmpl : Paper = {SCSSconf=>ECSS"SUSEXMPL 2015EUS"SCSS, title=>ECSS"SUSAn Example PaperEUS"}
-\end{lstlisting}
-%In either case, notice that we used literal syntax to introduce the component values, which are of regular string type.
-
-Given a value of type \lstinline{Paper}, we can project out a component value by providing  a  label:% like this:
-\begin{lstlisting}[numbers=none]
-let val exmpl_conf = # <SNATconfENAT> exmpl 
-\end{lstlisting}
-Here, \lstinline{#} identifies an \emph{operator constructor} (or \emph{opcon}) parameterized by a statically valued label, written literally here as \lstinline{<SNATconfENAT>}.
+% % Alternatively, we could have used the following construction to declare \lstinline{Paper} \emph{generatively}:
+% % \begin{lstlisting}[numbers=none]
+% % lprodtype Paper {
+% %   SPCTtitle : EPCTrstring /SURL.+EURL/
+% %    SPCTconf : EPCTrstring /SURL([A-Z]+) (\d\d\d\d)EURL/
+% % }
+% % \end{lstlisting}
+% % The difference is in how type equality is handled. Two labeled product types with the same signature are equal, unless declared generatively, in which case type equality is nominal.
+
+% Clients can introduce regular strings in analytic position using standard string literal syntax, as long as the body of the literal is in the corresponding regular language. 
+% Clients can introduce a value of type \lstinline{Paper} in an analytic position in one of two ways. We can omit the labels and provide component values positionally:
+% \begin{lstlisting}[numbers=none]
+% let val exmpl : Paper = {"SUSAn Example PaperEUS"SCSS, ECSS"SUSEXMPL 2015EUS"}
+% \end{lstlisting}
+% Alternatively, we can include the component labels explicitly for readability or if we want to give the components in an alternative order:
+% \begin{lstlisting}[numbers=none]
+% let val exmpl : Paper = {SCSSconf=>ECSS"SUSEXMPL 2015EUS"SCSS, title=>ECSS"SUSAn Example PaperEUS"}
+% \end{lstlisting}
+% %In either case, notice that we used literal syntax to introduce the component values, which are of regular string type.
+
+% Given a value of type \lstinline{Paper}, we can project out a component value by providing  a  label:% like this:
+% \begin{lstlisting}[numbers=none]
+% let val exmpl_conf = # <SNATconfENAT> exmpl 
+% \end{lstlisting}
+% Here, \lstinline{#} identifies an \emph{operator constructor} (or \emph{opcon}) parameterized by a statically valued label, written literally here as \lstinline{<SNATconfENAT>}.
 The resulting \emph{operator}, \lstinline{# <SNATconfENAT>}, is then passed one \emph{argument}, \lstinline{exmpl}.
@@
 
-We can project out the first captured group from the regular string \lstinline{exmpl_conf} using the \lstinline{#group} operator constructor, which is parameterized by a statically valued natural number referring to the group index:
-\begin{lstlisting}[numbers=none]
-let val exmpl_conf_name = #group 0 exmpl_conf
-\end{lstlisting}
-The variable \lstinline{exmpl_conf_name} has type \lstinline{rstring /SURL[A-Z]+EURL/}.
-
-We will equip labeled tuple types with an extended set of operators that go beyond the introduction and projection operators just discussed.
+% We can project out the first captured group from the regular string \lstinline{exmpl_conf} using the \lstinline{#group} operator constructor, which is parameterized by a statically valued natural number referring to the group index:
+% \begin{lstlisting}[numbers=none]
+% let val exmpl_conf_name = #group 0 exmpl_conf
+% \end{lstlisting}
+% The variable \lstinline{exmpl_conf_name} has type \lstinline{rstring /SURL[A-Z]+EURL/}.
+
+% We will equip labeled tuple types with an extended set of operators that go beyond the introduction and projection operators just discussed.
 For example, two labeled tuples can be concatenated (with common components updated with the value on the right) using the \lstinline{ltuple+} operator.
 Using this operator, we can add a component labeled \lstinline{authors} to \lstinline{exmpl}:
-\begin{lstlisting}[numbers=none]
-let val a : ltuple {SPCTauthors : EPCTlist(rstring /SURL.+EURL/)} = {["SUSHarry Q. BovikEUS"]}
-let val exmpl_paper_final = ltuple+ exmpl a
-\end{lstlisting}
-We can drop a component using the operator constructor \lstinline{ltuple-}, which is parameterized by a static label:
-\begin{lstlisting}[numbers=none]
-let val exmpl_paper_anon = ltuple- <SNATauthorsENAT> exmpl_paper_final
-\end{lstlisting}
-There are many other operators that we might wish to describe, for both labeled tuples \cite{Cardelli91operationson} and regular strings \cite{sanitation-psp14}, but for our purposes it suffices to stop here.
-
-\subsection{Existing Approaches}\label{sec:metamodules-related}
-Before describing how we express the type structure of labeled tuples and regular strings in Verse using metamodules, let us consider some alternative approaches for each.
-
-\subsubsection{Labeled Tuples}
-Recall that Verse builds only nullary and binary products into the IL, so we cannot express the type structure described above for labeled tuples  directly using IL primitives.
+% \begin{lstlisting}[numbers=none]
+% let val a : ltuple {SPCTauthors : EPCTlist(rstring /SURL.+EURL/)} = {["SUSHarry Q. BovikEUS"]}
+% let val exmpl_paper_final = ltuple+ exmpl a
+% \end{lstlisting}
+% We can drop a component using the operator constructor \lstinline{ltuple-}, which is parameterized by a static label:
+% \begin{lstlisting}[numbers=none]
+% let val exmpl_paper_anon = ltuple- <SNATauthorsENAT> exmpl_paper_final
+% \end{lstlisting}
+% There are many other operators that we might wish to describe, for both labeled tuples \cite{Cardelli91operationson} and regular strings \cite{sanitation-psp14}, but for our purposes it suffices to stop here.
+
+% \subsection{Existing Approaches}\label{sec:metamodules-related}
+% Before describing how we express the type structure of labeled tuples and regular strings in Verse using metamodules, let us consider some alternative approaches for each.
+
+% \subsubsection{Labeled Tuples}
+% Recall that Verse builds only nullary and binary products into the IL, so we cannot express the type structure described above for labeled tuples  directly using IL primitives.
 The Verse EL does not build in even these, but so as to separate concerns, let us assume for the moment that regular string types are available in the EL: %So to consider existing approaches, we have to at least somewhat weaken our insistence on minimalism.
-\begin{lstlisting}[numbers=none]
- (* type synonyms for concision *)
- let type title_t = rstring /SURL.+EURL/
- let type conf_t = rstring /SURL([A-Z]+) (\d\d\d\d)EURL/
-\end{lstlisting}
-
-\paragraph{Modules}
-It is reasonable to ask if it is possible to express the type structure of any particular labeled tuple type using the module system.
+% \begin{lstlisting}[numbers=none]
+%  (* type synonyms for concision *)
+%  let type title_t = rstring /SURL.+EURL/
+%  let type conf_t = rstring /SURL([A-Z]+) (\d\d\d\d)EURL/
+% \end{lstlisting}
+
+% \paragraph{Modules}
+% It is reasonable to ask if it is possible to express the type structure of any particular labeled tuple type using the module system.
 %For simplicity below, let us slightly weaken our insistence on minimalism and build binary products into the EL (and, for now, assume that regular string types are also built in in some suitable manner).%This is not strictly necessary -- we could use a Church-style encoding -- but this simplifies our examples.%then we could try to use the module system to express the type structure of any particular labeled tuple type.
 For our example above:
-For our example type \lstinline{Paper} from above, we might start by defining the following signature:
-\begin{lstlisting}
-signature PAPER = sig {
-  type t
-  (* introduction without labels *)
-  val intro : title_t -> conf_t -> t
-  (* introduction with explicit labels, in primary order *)
-  val intro_title_conf : title_t -> conf_t -> t
-  (* projection *)
-  val prj_title : t -> title_t
-  val prj_conf : t -> conf_t
-}
-\end{lstlisting}
-Here, we've taken all possible valid introductory and projection operators and called for their expression as functions.
+% For our example type \lstinline{Paper} from above, we might start by defining the following signature:
+% \begin{lstlisting}
+% signature PAPER = sig {
+%   type t
+%   (* introduction without labels *)
+%   val intro : title_t -> conf_t -> t
+%   (* introduction with explicit labels, in primary order *)
+%   val intro_title_conf : title_t -> conf_t -> t
+%   (* projection *)
+%   val prj_title : t -> title_t
+%   val prj_conf : t -> conf_t
+% }
+% \end{lstlisting}
+% Here, we've taken all possible valid introductory and projection operators and called for their expression as functions.
 Expressing labels explicitly is an important facet of having labeled tuple types in a language, so we put them into the function identifiers.
@@
 
-Even with our minimal EL, we could  implement the semantics of the corresponding operators against this signature using a Church-style encoding (the details are standard and omitted here).
+% Even with our minimal EL, we could  implement the semantics of the corresponding operators against this signature using a Church-style encoding (the details are standard and omitted here).
 Alternatively, if we slightly relax  our insistence on minimalism and add binary products to the EL, we could use them to implement our desired semantics like this:
-\begin{lstlisting}
- module Paper : PAPER = mod {
-   type t = title_t * conf_t
-   fun intro title conf => (title, conf)
-   fun intro_title_conf title conf => (title, conf)
-   fun prj_title (title, conf) => title 
-   fun prj_conf (title, conf) => conf
- }
- \end{lstlisting}
-%Compared to the four line type definition at the beginning of Sec.
+% \begin{lstlisting}
+%  module Paper : PAPER = mod {
+%    type t = title_t * conf_t
+%    fun intro title conf => (title, conf)
+%    fun intro_title_conf title conf => (title, conf)
+%    fun prj_title (title, conf) => title 
+%    fun prj_conf (title, conf) => conf
+%  }
+%  \end{lstlisting}
+% %Compared to the four line type definition at the beginning of Sec.
 safe to say that this is an  unacceptable departure from the four line type definition at the beginning of Sec.
@@
 
-There are several fundamental problems with this approach.
+% There are several fundamental problems with this approach.
 First, not every module matching the signature \lstinline{PAPER} correctly implements the semantics of the operators we seek to express, so each such module would itself need to be verified.
@@
 
-Even if we were to automate this (using, for example, \emph{type macros} as  in Scala \cite{ScalaMacros2013}, adapted to support Verse's modules), another issue is that we can only reasonably express the introductory and projection operators by specifying their semantics as functions like this.
+% Even if we were to automate this (using, for example, \emph{type macros} as  in Scala \cite{ScalaMacros2013}, adapted to support Verse's modules), another issue is that we can only reasonably express the introductory and projection operators by specifying their semantics as functions like this.
 To enumerate all possible operators that arise from the opcon \lstinline{ltuple-} using functions would require constructing an encoding of every labeled tuple type that might arise from dropping one or more components from the labeled tuple in question.
 If fully enumerated, this would add an  exponential factor to our volume of boilerplate code.
-Finally, there is simply no way to specify the semantics \lstinline{ltuple+} with a finite collection of functions because there are infinitely many extensions of any labeled tuple type.
+% Finally, there is simply no way to specify the semantics \lstinline{ltuple+} with a finite collection of functions because there are infinitely many extensions of any labeled tuple type.
 % of this operator would need to be implemented by clients manually, defeating the purpose of the operator.
@@
 
-% \begin{lstlisting}[numbers=none]
-% signature PAPER
-%   type t
-%   val intro_tpl : rstring /.../ * rstring /.../ -> t
-%   val intro_conf_title : rstring /.../ * rstring /.../ -> t
-%   val intro_title_conf : rstring /.../ * rstring /.../ -> t
-%   val prj_conf : t -> rstring /.../
-%   val prj_title : t -> rstring /.../
-%  \end{lstlisting}
-
-\paragraph{Records and Tuples}
-If, instead of this module-based encoding, we further weaken our insistence on minimalism and primitively build record and tuple types into the EL, as SML and many other languages have done, let us consider what might be possible.
+% % \begin{lstlisting}[numbers=none]
+% % signature PAPER
+% %   type t
+% %   val intro_tpl : rstring /.../ * rstring /.../ -> t
+% %   val intro_conf_title : rstring /.../ * rstring /.../ -> t
+% %   val intro_title_conf : rstring /.../ * rstring /.../ -> t
+% %   val prj_conf : t -> rstring /.../
+% %   val prj_title : t -> rstring /.../
+% %  \end{lstlisting}
+
+% \paragraph{Records and Tuples}
+% If, instead of this module-based encoding, we further weaken our insistence on minimalism and primitively build record and tuple types into the EL, as SML and many other languages have done, let us consider what might be possible.
 Note that these too are library constructs in Verse, but because they are so commonly built in to other languages, this situation is worth considering.%we can retreat from trying to use the module system, but have only slightly better options.
 
-The simplest approach we might consider taking is to directly map each labeled tuple type to a record type having an identically written component specification:
-\begin{lstlisting}[numbers=none]
-let type Paper_rcd = record {
-  title : title_t
-   conf : conf_t 
-}\end{lstlisting}
-Unlike labeled tuple types, record types are identified up to component reordering, so this mapping does not preserve certain type disequalities between labeled tuple types.
+% The simplest approach we might consider taking is to directly map each labeled tuple type to a record type having an identically written component specification:
+% \begin{lstlisting}[numbers=none]
+% let type Paper_rcd = record {
+%   title : title_t
+%    conf : conf_t 
+% }\end{lstlisting}
+% Unlike labeled tuple types, record types are identified up to component reordering, so this mapping does not preserve certain type disequalities between labeled tuple types.
 Without positional information available in the type, it is not  possible to allow clients to omit component labels when introducing a value of record type.%, because this information is not maintained by the type system.
 
-To work around this limitation if this is the only reason we care about preserving these type disequalities, we could define two conversion functions involving unlabeled tuples:
-\begin{lstlisting}[numbers=none]
-fun Paper_of_tpl (t : title_t, c : conf_t) => 
-  {title => t, conf => c}
-fun tpl_of_Paper {title => t, conf => c} =>
-  (t, c)
-\end{lstlisting}
-For clients to be able to rely on this approach, we must extrinsically verify that the library provider implemented these functions correctly.
+% To work around this limitation if this is the only reason we care about preserving these type disequalities, we could define two conversion functions involving unlabeled tuples:
+% \begin{lstlisting}[numbers=none]
+% fun Paper_of_tpl (t : title_t, c : conf_t) => 
+%   {title => t, conf => c}
+% fun tpl_of_Paper {title => t, conf => c} =>
+%   (t, c)
+% \end{lstlisting}
+% For clients to be able to rely on this approach, we must extrinsically verify that the library provider implemented these functions correctly.
 Clients must also bear the syntactic and dynamic costs of explicitly invoking these functions.
@@
 
-If, on the other hand, we do care about preserving all type disequalities between labeled tuple types, the workaround is even less elegant.
+% If, on the other hand, we do care about preserving all type disequalities between labeled tuple types, the workaround is even less elegant.
 We first need to define a record type with component labels tagged in some sufficiently unambiguous way by position, e.g.:
-\begin{lstlisting}[numbers=none]
-let type Paper = record {
-	__0_title : title_t
-	__1_conf : conf_t 
-}\end{lstlisting}
-To avoid exposing these labels directly to clients, we need two auxiliary type definitions:
-\begin{lstlisting}[numbers=none]
-let type Paper_tpl = (title_t * conf_t)
-let type Paper_rcd = (* as above *)
-\end{lstlisting}
-and four conversion functions:
-\begin{lstlisting}[numbers=none]
-fun Paper_of_tpl (t : title_t, c : conf_t) => {
-  __0_title => t,
-  __1_conf => c
-}
-fun tpl_of_Paper {__0_title => t, __1_conf => c} => 
-  (t, c)
-fun Paper_of_rcd {title => t, conf => c} => {
-  __0_title => t,
-  __1_conf => c
-}
-fun rcd_of_Paper {__0_title => t, __1_conf => c} = {
-  title => t,
-  conf => c
-}
-\end{lstlisting}
-Clients once again bear the costs of applying these conversion functions ahead of every operation and providers again bear the burden of defining this boilerplate code (which has volume linear in the number of components, albeit with a constant factor that is again not easy to dismiss) and verifying that these functions were implemented correctly.
+% \begin{lstlisting}[numbers=none]
+% let type Paper = record {
+% 	__0_title : title_t
+% 	__1_conf : conf_t 
+% }\end{lstlisting}
+% To avoid exposing these labels directly to clients, we need two auxiliary type definitions:
+% \begin{lstlisting}[numbers=none]
+% let type Paper_tpl = (title_t * conf_t)
+% let type Paper_rcd = (* as above *)
+% \end{lstlisting}
+% and four conversion functions:
+% \begin{lstlisting}[numbers=none]
+% fun Paper_of_tpl (t : title_t, c : conf_t) => {
+%   __0_title => t,
+%   __1_conf => c
+% }
+% fun tpl_of_Paper {__0_title => t, __1_conf => c} => 
+%   (t, c)
+% fun Paper_of_rcd {title => t, conf => c} => {
+%   __0_title => t,
+%   __1_conf => c
+% }
+% fun rcd_of_Paper {__0_title => t, __1_conf => c} = {
+%   title => t,
+%   conf => c
+% }
+% \end{lstlisting}
+% Clients once again bear the costs of applying these conversion functions ahead of every operation and providers again bear the burden of defining this boilerplate code (which has volume linear in the number of components, albeit with a constant factor that is again not easy to dismiss) and verifying that these functions were implemented correctly.
 %We are interested in all such pragmatic issues, so this is not an entirely satisfying solution.
@@
 
-As with the modular encoding of labeled tuples we attempted above, neither of these encodings using records and unlabeled tuples provides us with any way to uniformly express the more interesting operations, like \lstinline{ltuple+} or \lstinline{ltuple-}, that we discussed earlier.
-
-\subsubsection{Regular Strings}
-Let us now turn to regular string types. Note that we have specified the full semantics of regular string types in a recent workshop paper \cite{sanitation-psp14}.
-
-\paragraph{Dynamic Checks}
-The most common alternative to regular strings in existing languages is to simply use standard strings (which themselves may be expressed as lists or vectors of characters together with derived syntax) and insert dynamic checks around operations to maintain the regular string invariant.
+% As with the modular encoding of labeled tuples we attempted above, neither of these encodings using records and unlabeled tuples provides us with any way to uniformly express the more interesting operations, like \lstinline{ltuple+} or \lstinline{ltuple-}, that we discussed earlier.
+
+% \subsubsection{Regular Strings}
+% Let us now turn to regular string types. Note that we have specified the full semantics of regular string types in a recent workshop paper \cite{sanitation-psp14}.
+
+% \paragraph{Dynamic Checks}
+% The most common alternative to regular strings in existing languages is to simply use standard strings (which themselves may be expressed as lists or vectors of characters together with derived syntax) and insert dynamic checks around operations to maintain the regular string invariant.
 This clearly does not express the static semantics of regular strings, and incurs syntactic and dynamic cost.
 
-\paragraph{Type Refinements}
-We might attempt to recover the static semantics of regular strings by moving the logic governing regular string types into an external system of \emph{type refinements} \cite{Freeman91}.
+% \paragraph{Type Refinements}
+% We might attempt to recover the static semantics of regular strings by moving the logic governing regular string types into an external system of \emph{type refinements} \cite{Freeman91}.
 %This treats membership of a string in a specified regular language exclusively as a static verification condition.
-Such  systems allow programmers to specify \emph{refinements} of a type that capture more specific invariants.
+% Such  systems allow programmers to specify \emph{refinements} of a type that capture more specific invariants.
 For example, we might consider capturing the regular string invariant using refinements of the \lstinline{string} type.
@@
 In Verse, we might supply refinement specifications using structured comments:
-\begin{lstlisting}[numbers=none]
-let val conf : string (* ~ rstring /([A-Z]+) (\d\d\d\d)/ *) = "EXMPL 2015"
-\end{lstlisting}
-One immediate limitation of this approach is that there is now no way to express the group projection operator described above, i.e.
+% \begin{lstlisting}[numbers=none]
+% let val conf : string (* ~ rstring /([A-Z]+) (\d\d\d\d)/ *) = "EXMPL 2015"
+% \end{lstlisting}
+% One immediate limitation of this approach is that there is now no way to express the group projection operator described above, i.e.
 we cannot define a function \lstinline{#group} that can be applied like this:
-\begin{lstlisting}[numbers=none]
-let val conf_venue = #group 0 conf
-\end{lstlisting}
-The reason is that group projection does not have a function-like semantics -- it needs access to information that here is  available only in the type refinement of each regular string, i.e.
+% \begin{lstlisting}[numbers=none]
+% let val conf_venue = #group 0 conf
+% \end{lstlisting}
+% The reason is that group projection does not have a function-like semantics -- it needs access to information that here is  available only in the type refinement of each regular string, i.e.
 the definitions of the captured groups.
@@
 
-\begin{lstlisting}[numbers=none]
-let val conf_venue = nth 0 (match /SURL([A-Z]+) (\d\d\d\d)EURL/ conf)
-\end{lstlisting}
-This has higher syntactic and dynamic cost. Moreover, proving that there will not be an out-of-bounds error in expressions like this requires more sophisticated reasoning about dynamic behavior.
-%Refinement types are strictly a layer above the language, and thus cannot affect its dynamic semantics.
-
-Another consideration is that type refinements do not introduce type disequalities.
+% \begin{lstlisting}[numbers=none]
+% let val conf_venue = nth 0 (match /SURL([A-Z]+) (\d\d\d\d)EURL/ conf)
+% \end{lstlisting}
+% This has higher syntactic and dynamic cost. Moreover, proving that there will not be an out-of-bounds error in expressions like this requires more sophisticated reasoning about dynamic behavior.
+% %Refinement types are strictly a layer above the language, and thus cannot affect its dynamic semantics.
+
+% Another consideration is that type refinements do not introduce type disequalities.
 As a result, a compiler cannot use the invariants they capture to optimize the representation of a value.
@@
 
-%Put another way, type refinements are quite useful when one needs only to make ``behavioral'' distinctions within a type, but they are not appropriate when new type disequalities would be useful or new operators need to be introduced.
+% %Put another way, type refinements are quite useful when one needs only to make ``behavioral'' distinctions within a type, but they are not appropriate when new type disequalities would be useful or new operators need to be introduced.
 %Refinement types cannot be used for this purpose without the possibility of conflict.
-%This is also the reason why one cannot viably do away with primitive record types by claiming that they are refinements of finite mappings with labels as keys.
-
-%In summary, type refinements are appropriate when one needs only to make finer static distinctions within an existing type.
+% %This is also the reason why one cannot viably do away with primitive record types by claiming that they are refinements of finite mappings with labels as keys.
+
+% %In summary, type refinements are appropriate when one needs only to make finer static distinctions within an existing type.
 When one needs to introduce new operators or type disequalities that have ``downstream'' impact, refinement types are not the right solution.%Refinement types are useful when no non-trivial operators need to be introduced into the language and the refinement has no impact on representation, but the type system itself must be extended in some way in more advanced situations.
 
-% \paragraph{Modules}
-% To create type disequalities, we might consider again turning to the module system, defining a signature for each regular string type that we wish to work with.
+% % \paragraph{Modules}
+% % To create type disequalities, we might consider again turning to the module system, defining a signature for each regular string type that we wish to work with.
 Again, this requires a non-trivial amount of boilerplate code:
-% \begin{lstlisting}[numbers=none]
-% signature TITLE = sig {
-%   type t
-%   val intro : string -> t
-%   val strcase : t -> (unit -> 'a) -> (string * string -> 'a) -> 'a
-% }
-
-% signature CONF_GROUP_1 = sig {
-% 	type t
-% 	val intro : string -> t
-% 	val strcase : t -> (unit -> 'a) -> (sting * strng -> 'a) -> 'a 
-% }
-
-% signature CONF = sig {
-%   type t
-%   val intro : string -> t
-%   val strcase : t -> (unit -> 'a) -> (string * string -> 'a) -> 'a
-%   structure Grp0 : CONF_GROUP_0
-%   structure Grp1 : CONF_GROUP_1
-%   val prj_group_0 : t -> Grp0.t
-%   val prj_group_1 : t -> Grp1.t
-% }
-
-% \end{lstlisting}
-
-% As discussed previously, we do not have a guarantee that all implementations of these signatures correctly capture the semantics we seek to express, so they must each be verified externally.
+% % \begin{lstlisting}[numbers=none]
+% % signature TITLE = sig {
+% %   type t
+% %   val intro : string -> t
+% %   val strcase : t -> (unit -> 'a) -> (string * string -> 'a) -> 'a
+% % }
+
+% % signature CONF_GROUP_1 = sig {
+% % 	type t
+% % 	val intro : string -> t
+% % 	val strcase : t -> (unit -> 'a) -> (sting * strng -> 'a) -> 'a 
+% % }
+
+% % signature CONF = sig {
+% %   type t
+% %   val intro : string -> t
+% %   val strcase : t -> (unit -> 'a) -> (string * string -> 'a) -> 'a
+% %   structure Grp0 : CONF_GROUP_0
+% %   structure Grp1 : CONF_GROUP_1
+% %   val prj_group_0 : t -> Grp0.t
+% %   val prj_group_1 : t -> Grp1.t
+% % }
+
+% % \end{lstlisting}
+
+% % As discussed previously, we do not have a guarantee that all implementations of these signatures correctly capture the semantics we seek to express, so they must each be verified externally.
 Moreover, we must externally verify that every application  of the \lstinline{intro} operator is statically correct as well (i.e.
@@
 
-\subsection{Contributions}\label{sec:metamodules-solution}
-
-%An important point is that labeled products, as well as other ``record-like'' constructs, can  be expressed by a \emph{type-directed translation} targeting our IL (and indeed, many compilers take advantage of facts like this to simplify the number of cases that must be considered downstream).
+% \subsection{Contributions}\label{sec:metamodules-solution}
+
+% %An important point is that labeled products, as well as other ``record-like'' constructs, can  be expressed by a \emph{type-directed translation} targeting our IL (and indeed, many compilers take advantage of facts like this to simplify the number of cases that must be considered downstream).
 %it is inaccurate to state that they are merely syntactic sugar over such a language.
 %Object systems are perhaps even more diverse across languages\todo{citation}.
-%Other dialects are justified by a desire to proivde more special-purpose semantic constructs. For example, 
-%Substantially more sophisticated examples require at most only a very slightly more capable intermediate language (the IL of popular virtual machines, even those with flaws like ubiquitous \texttt{null} references, e.g.
+% %Other dialects are justified by a desire to proivde more special-purpose semantic constructs. For example, 
+% %Substantially more sophisticated examples require at most only a very slightly more capable intermediate language (the IL of popular virtual machines, even those with flaws like ubiquitous \texttt{null} references, e.g.
 JVM or CIL bytecode, are also sufficient if used strictly as translation targets).
 
-Verse introduces a \emph{metamodule system} that gives library providers more direct control over the type structure of the EL.
+% Verse introduces a \emph{metamodule system} that gives library providers more direct control over the type structure of the EL.
 Using the metamodule system, library providers can express all of the types and operators discussed above.
 
-Just as the Verse (i.e.
+% Just as the Verse (i.e.
 ML) module system is organized around \emph{signatures} and \emph{modules}, the Verse metamodule system is organized around \emph{metasignatures} and \emph{metamodules}.
@@
 We consider only labeled tuple types in this section (regular string types will follow analagously; we will discuss regular string types in more detail in the dissertation).
-%Briefly summarized:
-%\begin{itemize}
-%\item A \emph{metasignature} specifies a set of type and operator constructors (tycons and opcons). Each is parameterized by a specified \emph{kind} of \emph{static value}. 
-%\item A \emph{metamodule}  defines static functions that govern the typechecking and  translation to the IL of each type and operator constructor specified in the  metasignature. 
-%\end{itemize}
-
-%Static values are values of the \emph{typed static language} (SL), which itself forms a total typed lambda calculus, with \emph{static expressions} $\sigma$ classified by \emph{kinds} $\kappa$.
+% %Briefly summarized:
+% %\begin{itemize}
+% %\item A \emph{metasignature} specifies a set of type and operator constructors (tycons and opcons). Each is parameterized by a specified \emph{kind} of \emph{static value}. 
+% %\item A \emph{metamodule}  defines static functions that govern the typechecking and  translation to the IL of each type and operator constructor specified in the  metasignature. 
+% %\end{itemize}
+
+% %Static values are values of the \emph{typed static language} (SL), which itself forms a total typed lambda calculus, with \emph{static expressions} $\sigma$ classified by \emph{kinds} $\kappa$.
 Types are one kind of static value.
@@
 
-\subsubsection{Example: Labeled Tuple Types via Metamodules}
-
-Figure \ref{fig:ltuplex} defines a metasignature, \lstinline{LTUPLE}, and a matching 
-metamodule, \lstinline{Ltuple}.
+% \subsubsection{Example: Labeled Tuple Types via Metamodules}
+
+% Figure \ref{fig:ltuplex} defines a metasignature, \lstinline{LTUPLE}, and a matching 
+% metamodule, \lstinline{Ltuple}.
 It also defines a \emph{metamodule-parameterized TSM} \lstinline{ltpl}, which will give us the introductory syntax shown in Sec.
@@
 
-\paragraph{Type Constructors} On line \ref{line:ltuplex-c-sig} of Figure \ref{fig:ltuplex}, the metasignature \lstinline{LTUPLE} specifies that all matching metamodules must define a type constructor \lstinline{c} parameterized by {static values} of {sort} \lstinline{ComponentSpec}.
+% \paragraph{Type Constructors} On line \ref{line:ltuplex-c-sig} of Figure \ref{fig:ltuplex}, the metasignature \lstinline{LTUPLE} specifies that all matching metamodules must define a type constructor \lstinline{c} parameterized by {static values} of {sort} \lstinline{ComponentSpec}.
 %This user-defined sort (not shown) classifies ordered mappings from labels (of sort \lstinline{Lbl}, also user-defined but not shown) to types, which are static values of sort \lstinline{Ty} (a primitive sort).
-Let us assume that we can write static values of sort \lstinline{ComponentSpec} concretely like this (in the dissertation, we will discuss how it would be possible to adapt TSLs to also operate at the level of static values):
-\begin{lstlisting}[numbers=none]
-let static val Paper_tyidx :: ComponentSpec = {
-  SPCTtitle : EPCTrstring /SURL.+EURL/SPCT,EPCT
-   SPCTconf : EPCTrstring /SURL([A-Z]+) (\d\d\d\d)EURL/
-}
-\end{lstlisting}
-Applying the type constructor \lstinline{Ltuple.c} to this parameter gives us a type:
-\begin{lstlisting}[numbers=none]
-let type Paper = Ltuple.c Paper_tyidx
-\end{lstlisting}
-Line \ref{line:tycon-synonym} defines the tycon synonym \lstinline{ltuple} for \lstinline{Ltuple.c}.
+% Let us assume that we can write static values of sort \lstinline{ComponentSpec} concretely like this (in the dissertation, we will discuss how it would be possible to adapt TSLs to also operate at the level of static values):
+% \begin{lstlisting}[numbers=none]
+% let static val Paper_tyidx : ComponentSpec = {
+%   SPCTtitle : EPCTrstring /SURL.+EURL/SPCT,EPCT
+%    SPCTconf : EPCTrstring /SURL([A-Z]+) (\d\d\d\d)EURL/
+% }
+% \end{lstlisting}
+% Applying the type constructor \lstinline{Ltuple.c} to this parameter gives us a type:
+% \begin{lstlisting}[numbers=none]
+% let type Paper = Ltuple.c Paper_tyidx
+% \end{lstlisting}
+% Line \ref{line:tycon-synonym} defines the tycon synonym \lstinline{ltuple} for \lstinline{Ltuple.c}.
 Substituting into the above, we recover the definition of \lstinline{Paper} given at the beginning of Sec.
 \ref{sec:metamodules-example}:
-\begin{lstlisting}[numbers=none]
-let type Paper = ltuple {
-  SPCTtitle : EPCTrstring /SURL.+EURL/SPCT,
-   conf : EPCTrstring /SURL([A-Z]+) (\d\d\d\d)EURL/
-}
-\end{lstlisting}
-Note that because types are static values of sort \lstinline{Ty}, this is equivalent to writing:
-\begin{lstlisting}[numbers=none]
-let static val Paper :: Ty = ltuple {
-  SPCTtitle : EPCTrstring /SURL.+EURL/SPCT,
-   conf : EPCTrstring /SURL([A-Z]+) (\d\d\d\d)EURL/
-}
-\end{lstlisting}
-
-Also note that to ensure that type equality is decidable, tycon parameters can only be of a sort for which equality coincides with structural comparison of values.
+% \begin{lstlisting}[numbers=none]
+% let type Paper = ltuple {
+%   SPCTtitle : EPCTrstring /SURL.+EURL/SPCT,
+%    conf : EPCTrstring /SURL([A-Z]+) (\d\d\d\d)EURL/
+% }
+% \end{lstlisting}
+% Note that because types are static values of sort \lstinline{Ty}, this is equivalent to writing:
+% \begin{lstlisting}[numbers=none]
+% let static val Paper : Ty = ltuple {
+%   SPCTtitle : EPCTrstring /SURL.+EURL/SPCT,
+%    conf : EPCTrstring /SURL([A-Z]+) (\d\d\d\d)EURL/
+% }
+% \end{lstlisting}
+
+% Also note that to ensure that type equality is decidable, tycon parameters can only be of a sort for which equality coincides with structural comparison of values.
 We call these sorts \emph{equality sorts}.
@@
 
-\paragraph{Type Translations} Recall from Sec.
+% \paragraph{Type Translations} Recall from Sec.
 \ref{sec:verse} that each external type must translate to an internal type.
@@
 
-The  type translator for \lstinline{Ltuple.c} is defined on lines \ref{line:c-start}-\ref{line:c-end} of Figure \ref{fig:ltuplex}.
+% The  type translator for \lstinline{Ltuple.c} is defined on lines \ref{line:c-start}-\ref{line:c-end} of Figure \ref{fig:ltuplex}.
 We have chosen here to translate labeled tuple types to nested binary product types internally.
-The type translator generates these by folding over the component mapping (using helper function \lstinline{comp_spec_to_tuples :: ComponentSpec -> list(Lbl * Ty)}, not shown).
+The type translator generates these by folding over the component mapping (using helper function \lstinline{comp_spec_to_tuples : ComponentSpec -> list(Lbl * Ty)}, not shown).
 It generates the representation of \lstinline{unit} for the empty case, and nests the binary product types in the recursive case (using the standard \emph{unquote} syntax, \lstinline{%}).
-For example, the representation of the type translation of \lstinline{Paper} determined by this static function is:
-\begin{lstlisting}[numbers=none]
-`SCSStrans(ECSSrstring /SURL.+EURL/SCSS) * (trans(ECSSrstring /SURL([A-Z]+) (\d\d\d\d)EURL/SCSS) * unit)ECSS`
-\end{lstlisting}
-Notice that the translations of the component regular string types are not inserted directly, but are rather treated {parametrically} using the \emph{translational form} \lstinline{SCSStrans(ECSSTSCSS)}.
+% For example, the representation of the type translation of \lstinline{Paper} determined by this static function is:
+% \begin{lstlisting}[numbers=none]
+% `SCSStrans(ECSSrstring /SURL.+EURL/SCSS) * (trans(ECSSrstring /SURL([A-Z]+) (\d\d\d\d)EURL/SCSS) * unit)ECSS`
+% \end{lstlisting}
+% Notice that the translations of the component regular string types are not inserted directly, but are rather treated {parametrically} using the \emph{translational form} \lstinline{SCSStrans(ECSSTSCSS)}.
 This will be key to the modularity principle -- \emph{translation independence} -- that we will discuss shortly.%\todo{discuss SL}
 
-Let us assume, for example, that all regular string types translate to standard internal strings, encoded by some internal type abbreviated \lstinline{string}.
+% Let us assume, for example, that all regular string types translate to standard internal strings, encoded by some internal type abbreviated \lstinline{string}.
 After we substitute this in for these translational forms, we arrive at the translation of \lstinline{Paper}:
-\begin{lstlisting}[numbers=none]
-string * (string * unit)
-\end{lstlisting}
-
-Formally, we should be able to derive that $\vdash \texttt{Paper}~\mathtt{type} \leadsto \texttt{string} \times (\texttt{string} \times \mathsf{unit})$. The following rule governs type translation:
-\begin{mathpar}
-\inferrule[trans-ty]{
-    \vdash_\Phi \mathbf{ty}(m\cdot c; \sigma_\text{param}) :: \mathsf{Ty}\\
-    \mathbf{ty}(m\cdot c; \sigma_\text{param})~\mathtt{sval}_\Phi\\
-	\mathbf{translator}(\Phi; m\cdot c)=\sigma_\text{translator}\\
-	\sigma_\text{translator}(\sigma_\text{param}) \Downarrow \sigma_\text{trans}\\
-	\sigma_\text{trans} \uparrow_\Phi^m \tau_\text{abstrans} \parallel {\delta} : {\Delta}\\
-	{\Delta} \vdash \tau_\text{abstrans}~\mathtt{itype}
-}{
-	\vdash_\Phi \mathbf{ty}(m\cdot c;\sigma_\text{param})~\mathtt{type}\leadsto[{\delta}]\tau_\text{abstrans}
-}
-\end{mathpar}
-Note that this rule is again slightly simplified here.
+% \begin{lstlisting}[numbers=none]
+% string * (string * unit)
+% \end{lstlisting}
+
+% Formally, we should be able to derive that $\vdash \texttt{Paper}~\mathtt{type} \leadsto \texttt{string} \times (\texttt{string} \times \mathsf{unit})$. The following rule governs type translation:
+% \begin{mathpar}
+% \inferrule[trans-ty]{
+%     \vdash_\Phi \mathbf{ty}(m\cdot c; \sigma_\text{param}) : \mathsf{Ty}\\
+%     \mathbf{ty}(m\cdot c; \sigma_\text{param})~\mathtt{sval}_\Phi\\
+% 	\mathbf{translator}(\Phi; m\cdot c)=\sigma_\text{translator}\\
+% 	\sigma_\text{translator}(\sigma_\text{param}) \Downarrow \sigma_\text{trans}\\
+% 	\sigma_\text{trans} \uparrow_\Phi^m \tau_\text{abstrans} \parallel {\delta} : {\Delta}\\
+% 	{\Delta} \vdash \tau_\text{abstrans}~\mathtt{itype}
+% }{
+% 	\vdash_\Phi \mathbf{ty}(m\cdot c;\sigma_\text{param})~\mathtt{type}\leadsto[{\delta}]\tau_\text{abstrans}
+% }
+% \end{mathpar}
+% Note that this rule is again slightly simplified here.
 It would need to be modified to handle external type abstraction, for example.
@@
 
-Looking at the conclusion of the rule, we see that in the abstract syntax of the SL, types constructed by user-defined tycons take the form $\textbf{ty}(m\cdot c; \sigma_\text{param})$, where $m\cdot c$ refers to a  {tycon} $c$ defined by metamodule $m$ and $\sigma_\text{param}$ is the type parameter.
+% Looking at the conclusion of the rule, we see that in the abstract syntax of the SL, types constructed by user-defined tycons take the form $\textbf{ty}(m\cdot c; \sigma_\text{param})$, where $m\cdot c$ refers to a  {tycon} $c$ defined by metamodule $m$ and $\sigma_\text{param}$ is the type parameter.
 The premises can be understood as follows:
-\begin{enumerate}
-\item The first premise checks that the type is of the required sort, $\mathsf{Ty}$.
+% \begin{enumerate}
+% \item The first premise checks that the type is of the required sort, $\mathsf{Ty}$.
 This involves finding the parameter sort associated with the tycon in the \emph{metamodule context}, $\Phi$ (we omit the straightforward definitions and rules for concision here).
-\item The second premise checks that the parameter is a static value. %All static expressions, including types, are fully evaluated during typechecking and translation.
-\item The third premise looks up the translator associated with $m \cdot c$ in $\Phi$. 
-\item The fourth premise applies this translator to the type parameter to generate the representation of the type translation, $\sigma_\text{trans}$ (e.g. shown above for \lstinline{Paper}).
-\item The fifth premise decodes $\sigma_\text{trans}$ to produce an \emph{abstracted type translation}, $\tau_\text{abstrans}$.
+% \item The second premise checks that the parameter is a static value. %All static expressions, including types, are fully evaluated during typechecking and translation.
+% \item The third premise looks up the translator associated with $m \cdot c$ in $\Phi$. 
+% \item The fourth premise applies this translator to the type parameter to generate the representation of the type translation, $\sigma_\text{trans}$ (e.g. shown above for \lstinline{Paper}).
+% \item The fifth premise decodes $\sigma_\text{trans}$ to produce an \emph{abstracted type translation}, $\tau_\text{abstrans}$.
 It is ``abstracted'' in that translational forms referring to types constructed by metamodules other than $m$, e.g.
@@
 The context corresponding to this substitution, here $\Delta=\alpha_1, \alpha_2$, is also generated.
-\item The final premise validates the abstracted type translation under $\Delta$ by making sure it is a valid internal type according to the internal statics.
-\item Finally, the conclusion of the rule applies $\delta$ to the abstracted translation to produce the final type translation.
-\end{enumerate}
-
-We will see why having this ability to selectively hold type translations abstract is critical to modularity after we consider operator constructors.
-
-\paragraph{Operator Constructors} On line \ref{line:intro-unlabeled-sig} of Figure \ref{fig:ltuplex}, the metasignature \lstinline{LTUPLE} specifies an operator constructor, \lstinline{intro_unlabeled}, parameterized by static values of sort \lstinline{unit}.
+% \item The final premise validates the abstracted type translation under $\Delta$ by making sure it is a valid internal type according to the internal statics.
+% \item Finally, the conclusion of the rule applies $\delta$ to the abstracted translation to produce the final type translation.
+% \end{enumerate}
+
+% We will see why having this ability to selectively hold type translations abstract is critical to modularity after we consider operator constructors.
+
+% \paragraph{Operator Constructors} On line \ref{line:intro-unlabeled-sig} of Figure \ref{fig:ltuplex}, the metasignature \lstinline{LTUPLE} specifies an operator constructor, \lstinline{intro_unlabeled}, parameterized by static values of sort \lstinline{unit}.
 The qualifier \lstinline{ana} specifies that this opcon is only for use in analytic positions (i.e.
@@
 For example, we can apply \lstinline{Ltuple.intro_unlabeled} as follows, because the return type of the function determines the expected type:
-\begin{lstlisting}[numbers=none]
-fun make_paper(c : conf_t, t : title_t) : Paper => 
-  Ltuple.intro_unlabeled () (c; t)
-\end{lstlisting}
-First, we provided the opcon with a static parameter value of the appropriate sort, here \lstinline{unit}, to create an operator.
+% \begin{lstlisting}[numbers=none]
+% fun make_paper(c : conf_t, t : title_t) : Paper => 
+%   Ltuple.intro_unlabeled () (c; t)
+% \end{lstlisting}
+% First, we provided the opcon with a static parameter value of the appropriate sort, here \lstinline{unit}, to create an operator.
 Then, we supply the operator with an \emph{argument list}, \lstinline{(conf; title)}.
 
-This is syntactically unwieldy.
+% This is syntactically unwieldy.
 Luckily, we can define a parameterized TSM, \lstinline{ltpl}, that gives us a more conventional syntax.
@@
 We elide the parser implementation, but as the comment indicates, we can rewrite \lstinline{make_paper} like this:
-\begin{lstlisting}[numbers=none]
-fun make_paper_tsm(c : conf_t, t : title_t) : Paper => 
-  ltpl Ltuple Paper_tyidx {cSCSS, ECSSt}
-\end{lstlisting}
-
-To avoid having to explicitly name the TSM and supply the parameters, we designate \lstinline{ltpl} as \lstinline{Ltuple.c}'s TSL on line \ref{line:ltplx-annotation}.
+% \begin{lstlisting}[numbers=none]
+% fun make_paper_tsm(c : conf_t, t : title_t) : Paper => 
+%   ltpl Ltuple Paper_tyidx {cSCSS, ECSSt}
+% \end{lstlisting}
+
+% To avoid having to explicitly name the TSM and supply the parameters, we designate \lstinline{ltpl} as \lstinline{Ltuple.c}'s TSL on line \ref{line:ltplx-annotation}.
 Now we can rewrite \lstinline{make_paper} in the standard way:
-\begin{lstlisting}[numbers=none]
-fun make_paper_tsl(c : conf_t, t : title_t) : Paper => 
-  {cSCSS, ECSSt}
-\end{lstlisting}
-
-Putting syntax aside, the metamodule defining the opcon, \lstinline{Ltuple}, programmatically determines how to typecheck and translate this operation, again by the defining a static function called the \emph{opcon definition}.
+% \begin{lstlisting}[numbers=none]
+% fun make_paper_tsl(c : conf_t, t : title_t) : Paper => 
+%   {cSCSS, ECSSt}
+% \end{lstlisting}
+
+% Putting syntax aside, the metamodule defining the opcon, \lstinline{Ltuple}, programmatically determines how to typecheck and translate this operation, again by the defining a static function called the \emph{opcon definition}.
 The definition of \lstinline{intro_unlabeled} is shown on lines \ref{line:introul-start}-\ref{line:introul-end} of Figure \ref{fig:ltuplex}.
@@
 
-On lines \ref{line:introul-start}-\ref{line:introul-end}, the opcon definition first checks that the type is actually a labeled tuple type using the \lstinline{tycase} primitive, which case analyzes against the type constructor of the provided type.
+% On lines \ref{line:introul-start}-\ref{line:introul-end}, the opcon definition first checks that the type is actually a labeled tuple type using the \lstinline{tycase} primitive, which case analyzes against the type constructor of the provided type.
 There must always be a default case to ensure exhaustiveness.
@@
 For our example above, the representation of the translation generated by the opcon definition is:
-\begin{lstlisting}[numbers=none]
-`SQT(anatrans[0](EQTconf_tSQT), (anatrans[1](EQTtitle_tSQT), ()))`
-\end{lstlisting}
-An \emph{argument translation} of the form \lstinline{SQTanatrans[i](EQTTSQT)} stands in for the translation of argument \lstinline{i} analyzed against type \lstinline{T}.
-
-The semantics next generates an \emph{abstracted translation}, which is a corresponding internal term where each argument translation has been replaced by a corresponding fresh variable.
+% \begin{lstlisting}[numbers=none]
+% `SQT(anatrans[0](EQTconf_tSQT), (anatrans[1](EQTtitle_tSQT), ()))`
+% \end{lstlisting}
+% An \emph{argument translation} of the form \lstinline{SQTanatrans[i](EQTTSQT)} stands in for the translation of argument \lstinline{i} analyzed against type \lstinline{T}.
+
+% The semantics next generates an \emph{abstracted translation}, which is a corresponding internal term where each argument translation has been replaced by a corresponding fresh variable.
 To validate it, we assume that the type of each variable is the abstracted type translation of the type the corresponding argument was analyzed against.
@@
 
-By treating the arguments to the operation parametrically in this way, we can be sure that the metamodule defining labeled tuple types cannot violate the translation invariants that other metamodules maintain.
+% By treating the arguments to the operation parametrically in this way, we can be sure that the metamodule defining labeled tuple types cannot violate the translation invariants that other metamodules maintain.
 For example, even if regular strings are implemented (by some other metamodule, not shown) as strings satisfying the regular string invariant, \lstinline{Ltuple} knows only that \emph{there exists} some translation for each regular string type, but nothing more.
@@
 
-\begin{lstlisting}[numbers=none]
-syn opcon bad {
-	static fn (op_param :: unit, args :: list(Arg)) :: Ty * IExp => 
-	  (rstring /.+/, `""`)
-}
-\end{lstlisting}
-Here, the opcon ignores the arguments and synthesizes a regular string type that the client will assume classifies non-empty strings.
+% \begin{lstlisting}[numbers=none]
+% syn opcon bad {
+% 	static fn (op_param : unit, args : list(Arg)) : Ty * IExp => 
+% 	  (rstring /.+/, `""`)
+% }
+% \end{lstlisting}
+% Here, the opcon ignores the arguments and synthesizes a regular string type that the client will assume classifies non-empty strings.
 However the corresponding translation is in fact the empty string.
@@
 
-%If the regular string metamodule later decides to change how it implements regular string types, this would require only local changes to that metamodule.
+% %If the regular string metamodule later decides to change how it implements regular string types, this would require only local changes to that metamodule.
 We call this property \emph{translation independence}, by analogy to the \emph{representation independence} property that underlies modular reasoning with ML modules.
@@
 
-To formalize the intuition just described, we will need to introduce a non-trivial number of technical devices.
+% To formalize the intuition just described, we will need to introduce a non-trivial number of technical devices.
 It would thus require too much space to give a comprehensive formal account of this mechanism in this proposal, so  we leave this as work that remains to be done (though see below for progress).
@@
 
-Note that the static functions that define these opcons contain code that looks just like the code that would appear in a standard implementation of the first stage of a compiler for a language that built in labeled tuple types.
+% Note that the static functions that define these opcons contain code that looks just like the code that would appear in a standard implementation of the first stage of a compiler for a language that built in labeled tuple types.
 The novelty is not in these definitions, but in how they are organized and brought into the language.
-% \begin{mathpar}
-% \inferrule[ap-anaop]{
-%   \mathbf{ana opcon}~m.\text{op}~\mathbf{of}~\uppi_\text{param}~\{ \sigma_\text{opdef} \} \in \Phi\\
-%   \vdash_\Phi \sigma_\text{param} :: \uppi_\text{param}\\
-%   \mathbf{args}(\es) = \sigma_\text{args}\\
-%   \sigma_\text{opdef} (\sigma_\text{ty}) (\sigma_\text{param}) (\sigma_\text{args}) \Downarrow_{\hat{\Gamma};\Phi;\es} \sigma_\text{trans}\\
-%   \sigma_\text{trans} \parallel \emptyset~\emptyset \uparrow_{\hat{\Gamma}; \Phi; \es}^m \tau_\text{abstrans} \parallel \memD~\memG\\
-%   \mathsf{trans}(\sigma_\text{ty}) \parallel \memD \uparrow_{\Phi}^m 
-% }{
-% 	\hat{\Gamma} \vdash_\Phi \mathbf{opconap}(m; \text{op}; \sigma_\text{param}; \dot{e}) \Leftarrow \sigma_\text{ty} \leadsto [\delta][\gamma]\iota
-% }
-% \end{mathpar}
-% omega anaop of pi_param { sigma_opdef }
-% sigma_param :: pi_param
-% arginterface(es) = sigma_args
-% sigma_opdef (sigma_ty) (sigma_param) (sigma_args) |v sigma_trans
-% sigma_trans || 0 0 ^| iota_trans |\ (g : G) (d : D)
-% trans(sigma) || d : D ^| tau_tytrans || d' : D'
-% D G |- iota_trans : tau_tytrans
-% -----------------------------------------------------------------
-% opconap(omega; sigma_param; es) <= sigma ~> [delta][gamma]iota
-
-\subsubsection{Limitations}
-The main limitation of the mechanism just described is that it only supports defining types and operators that do not require adding new contexts to the EL typing judgement.
+% % \begin{mathpar}
+% % \inferrule[ap-anaop]{
+% %   \mathbf{ana opcon}~m.\text{op}~\mathbf{of}~\uppi_\text{param}~\{ \sigma_\text{opdef} \} \in \Phi\\
+% %   \vdash_\Phi \sigma_\text{param} : \uppi_\text{param}\\
+% %   \mathbf{args}(\es) = \sigma_\text{args}\\
+% %   \sigma_\text{opdef} (\sigma_\text{ty}) (\sigma_\text{param}) (\sigma_\text{args}) \Downarrow_{\hat{\Gamma};\Phi;\es} \sigma_\text{trans}\\
+% %   \sigma_\text{trans} \parallel \emptyset~\emptyset \uparrow_{\hat{\Gamma}; \Phi; \es}^m \tau_\text{abstrans} \parallel \memD~\memG\\
+% %   \mathsf{trans}(\sigma_\text{ty}) \parallel \memD \uparrow_{\Phi}^m 
+% % }{
+% % 	\hat{\Gamma} \vdash_\Phi \mathbf{opconap}(m; \text{op}; \sigma_\text{param}; \dot{e}) \Leftarrow \sigma_\text{ty} \leadsto [\delta][\gamma]\iota
+% % }
+% % \end{mathpar}
+% % omega anaop of pi_param { sigma_opdef }
+% % sigma_param : pi_param
+% % arginterface(es) = sigma_args
+% % sigma_opdef (sigma_ty) (sigma_param) (sigma_args) |v sigma_trans
+% % sigma_trans || 0 0 ^| iota_trans |\ (g : G) (d : D)
+% % trans(sigma) || d : D ^| tau_tytrans || d' : D'
+% % D G |- iota_trans : tau_tytrans
+% % -----------------------------------------------------------------
+% % opconap(omega; sigma_param; es) <= sigma ~> [delta][gamma]iota
+
+% \subsubsection{Limitations}
+% The main limitation of the mechanism just described is that it only supports defining types and operators that do not require adding new contexts to the EL typing judgement.
 No new binding structure can be defined.
@@
 
-Establishing the metatheory of metamodules is also somewhat tricky, again because it involves reasoning about expressions that are not clearly subexpressions of those in the conclusion of the relevant rules.
+% Establishing the metatheory of metamodules is also somewhat tricky, again because it involves reasoning about expressions that are not clearly subexpressions of those in the conclusion of the relevant rules.
 We plan to give an induction principle by which we can prove the necessary metatheorems in the dissertation.
 
-\subsection{Timeline and Milestones}\label{sec:metamodules-timeline}
-The work above has recently been refactored so as to more closely follow the construction of the ML module system.
+% \subsection{Timeline and Milestones}\label{sec:metamodules-timeline}
+% The work above has recently been refactored so as to more closely follow the construction of the ML module system.
 Previously, the main organizational unit was a single tycon, rather than a metamodule.
@@
 They can be found at:
-\begin{itemize}
-\item Paper:\\ \url{https://github.com/cyrus-/papers/blob/master/atlam-icfp15/atlam-icfp15.pdf}
-\item Technical Report:\\ \url{https://github.com/cyrus-/papers/blob/master/atlam-icfp15/atlam-icfp15-supplement.pdf}
-\end{itemize}
-
-We propose the following concrete milestones, to be completed after the work in the previous section:
-\begin{enumerate}
-\item Refactor the syntax around metamodules \textbf{(1 week)}
-\item Update the static language's semantics \textbf{(1 week)}
-\item Update the external language's semantics \textbf{(1 week)}
-\item Update and complete the metatheory \textbf{(3 weeks)}
-\item Add support for external type abstraction \textbf{(1 week)}
-\end{enumerate}
-
-To further validate our work, we plan to detail a number of the examples mentioned above, including regular strings, datatypes, and a simple prototypic object system.
+% \begin{itemize}
+% \item Paper:\\ \url{https://github.com/cyrus-/papers/blob/master/atlam-icfp15/atlam-icfp15.pdf}
+% \item Technical Report:\\ \url{https://github.com/cyrus-/papers/blob/master/atlam-icfp15/atlam-icfp15-supplement.pdf}
+% \end{itemize}
+
+% We propose the following concrete milestones, to be completed after the work in the previous section:
+% \begin{enumerate}
+% \item Refactor the syntax around metamodules \textbf{(1 week)}
+% \item Update the static language's semantics \textbf{(1 week)}
+% \item Update the external language's semantics \textbf{(1 week)}
+% \item Update and complete the metatheory \textbf{(3 weeks)}
+% \item Add support for external type abstraction \textbf{(1 week)}
+% \end{enumerate}
+
+% To further validate our work, we plan to detail a number of the examples mentioned above, including regular strings, datatypes, and a simple prototypic object system.
 We anticipate that the writing involved to do this, along with other writing tasks, will take another \textbf{2 weeks}.
 So the total for this section is 9 weeks.
-% The semantics uses internal type abstraction to guarantee that metamodule providers can reason modularly about translation invariants, just as it uses external type abstraction to guarantee that module providers can reason modularly about representation invariants.%For example, a metamodule that expresses the type structure of records wouldLblTyList introduce a type constructor  \texttt{record} parameterized by a static mapping from labels to types, and two operators: the introductory operator \texttt{intro}, parameterized by a list of row labels, and the row projection operator \texttt{prj}, parameterized by a row label.
+% % The semantics uses internal type abstraction to guarantee that metamodule providers can reason modularly about translation invariants, just as it uses external type abstraction to guarantee that module providers can reason modularly about representation invariants.%For example, a metamodule that expresses the type structure of records wouldLblTyList introduce a type constructor  \texttt{record} parameterized by a static mapping from labels to types, and two operators: the introductory operator \texttt{intro}, parameterized by a list of row labels, and the row projection operator \texttt{prj}, parameterized by a row label.
 It might then choose to internally represent records by binary products, which the IL provides.
@@
 
-\section{Conclusion}
-In summary, we have proposed a new functional programming language, Verse, that gives programmers the ability to programmatically implement new syntactic expansions and new type structure using statically evaluated functions.
-Consequently, Verse does not need to build in constructs that comparable languages need to (or would need to) build in, like regular expression syntax, list syntax, labeled tuples and regular strings.
-These and other examples that we will cover in the dissertation will serve to validate our claim that Verse is highly expressive (and thus should lead to a decreased need for dialect formation).
+\section{Conclusion}\label{sec:conclusion}
+
+In summary, we have proposed a new sort of macro that gives programmers the ability to programmatically express new syntactic expansions using statically evaluated functions.
+These expansions are hygienic and restricted  to a single parameterized family of types, so that programmers can more easily read and reason about programs that use them.
+By using local type inference, the syntactic cost of using these syntactic expansions is kept low.
+Syntactic conflicts are impossible by construction.
+These macros also interact well with features found in typed functional languages like ML, including datatypes, modules and pattern matching.
+Consequently, Verse does not need to build in syntactic constructs that comparable languages need to (or would need to) build in, like regular expression syntax, list syntax and HTML syntax.
+% These and other examples that we will cover in the dissertation will serve to validate our claim that Verse is highly expressive (and thus should lead to a decreased need for dialect formation).
 
@@
 A more detailed specification and metatheoretic results that validate our claims that Verse has a hygienic type discipline and strong modular reasoning principles represent the main avenues for remaining work.
-We have already made substantial progress on closely related variants of the mechanisms proposed here, so we anticipate that the work can be completed over the course of 17 weeks, per the timelines in the sections above.
-Given that the proposal should be complete by the end of Summer 2015, this implies that the remaining work can be completed over the course of Fall 2015.
-Final updates and the presentation of the dissertation should then be in early Spring 2016.
+We have already made substantial progress on closely related variants of the mechanisms proposed here, so we anticipate that the work can be completed over the remainder of the semester, per the timeline above.
 
@@
 % \textbf{Sort} & & & \textbf{Abstract Form} & \textbf{Typeset Form} & \textbf{Description}\\
-% \mathsf{Type} & \tau & ::= & t & t & \text{type variable}\\
+% \mathsf{Type} & \tau & := & t & t & \text{type variable}\\
 % & & & \mathtt{arr}(\tau; \tau) & \tau \rightarrow \tau & \text{function type}\\
@@
 % \textbf{Sort} & & & \textbf{Abstract Form} & \textbf{Typeset Form} & \textbf{Description}\\
-% \mathsf{Exp} & \iota & ::= & x & x & \text{value variable}\\
+% \mathsf{Exp} & \iota & := & x & x & \text{value variable}\\
 % & & & \mathtt{lam}[\tau](x.\iota) & \lambda x{:}\tau.\iota & \text{value abstraction}\\
@@
 
-\begin{figure}[p!]
-\vspace{-25px}
-\begin{lstlisting}
-metasignature LTUPLE = metasig {
-  ^\label{line:ltuplex-c-sig}^tycon c of ComponentSpec
-  ^\label{line:intro-unlabeled-sig}^ana opcon intro_unlabeled of unit
-  ^\label{line:intro-labeled-sig}^syn opcon intro_labeled of list(Lbl)
-  ^\label{line:prj-sig}^syn opcon # of Lbl
-  ^\label{line:conc-sig}^syn opcon + of unit
-  ^\label{line:drop-sig}^syn opcon - of Lbl
-}
-
-^\label{line:ltplx-tsm-start}^syntax ltpl(L :: LTUPLE, param :: ComponentSpec) at L.c(param) {
-  static fn ps => (* 
-    - elaborates {e_1, ..., e_n} to 
-        L.intro_unlabeled () (e_1; ...; e_n)
-    - elaborates {lbl1 => e_1, ..., lbln => e_n} to 
-        L.intro_labeled [lbl1, ..., lbln] e_1 ... e_n
-  *)
-}^\label{ltplx-tsm-end}^
-
-metamodule Ltuple :: LTUPLE = metamod {
-  ^\label{line:c-start}^(* translate labeled tuple types to nested products *)
-  tycon c {
-  	(* type translation generator: *)
-  	static fn (param :: ComponentSpec) :: ITy => fold (comp_spec_to_tuples param)
-      `SCSSunitECSS`
-      (fn ((lbl, ty), cur_ty_trans) => `SCSStrans(ECSStySCSS) * %ECSScur_ty_trans`)
-  }^\label{line:c-end}^
-
-  ^\label{line:introul-start}^ana opcon intro_unlabeled {
-  	(* translate intro operator to nested pairs *)
-  	static fn (ty :: Ty, op_param :: unit, args :: list(Arg)) :: IExp => tycase(ty) {
-        c(ty_param) => (fold (zipExact (comp_spec_to_tuples ty_param) args)
-      	  `SCSS()ECSS`
-          (fn (((lbl, ty), arg), cur_trans) => `SCSS(%{ECSSana(arg, ty)SCSS}, %ECSScur_transSCSS)ECSS`)
-        )
-      | _ => raise TyErr("...")
-    }
-  }^\label{line:introul-end}^
-
-  ^\label{line:introl-start}^syn opcon intro_labeled {
-  	static fn (op_param :: list(Lbl), args :: list(Arg)) :: Ty * IExp => 
-      (* ... similar to intro_unlabeled but reorder first ... *)
-  }^\label{line:introl-end}^
-
-  ^\label{line:prj-start}^syn opcon # {
-  	static fn (op_param :: Lbl, args :: list(Arg)) :: Ty * IExp =>  
-      (* ... generate the appropriate n-fold projection ... *)
-  }^\label{line:prj-end}^
-
-  ^\label{line:ltpl-conc-start}^syn opcon + {
-  	static fn (op_param :: unit, args :: list(Arg)) :: Ty * IExp => 
-      (* generate the new type and combine the two nested tuples *)
-  }^\label{line:ltpl-conc-end}^
-
-  ^\label{line:ltpl-drop-start}^syn opcon - {
-  	static fn (op_param :: Lbl, args :: list(Arg)) :: Ty * IExp => 
-      (* generate the new type and drop the appropriate component *)
-  }^\label{line:ltpl-drop-end}^
-} with syntax ltpl^\label{line:ltplx-annotation}^
-(* synonyms used in Sec. 5.1 *)
-^\label{line:tycon-synonym}^let tycon ltuple = Ltuple.c
-let opcon # = Ltuple.#
-let opcon ltuple+ = Ltuple.+ 
-let opcon ltuple- = Ltuple.-
-\end{lstlisting}
-\caption{Using metamodules to define the type structure of labeled tuple types.}
-\label{fig:ltuplex}
-\end{figure}
+% \begin{figure}[p!]
+% \vspace{-25px}
+% \begin{lstlisting}
+% metasignature LTUPLE = metasig {
+%   ^\label{line:ltuplex-c-sig}^tycon c of ComponentSpec
+%   ^\label{line:intro-unlabeled-sig}^ana opcon intro_unlabeled of unit
+%   ^\label{line:intro-labeled-sig}^syn opcon intro_labeled of list(Lbl)
+%   ^\label{line:prj-sig}^syn opcon # of Lbl
+%   ^\label{line:conc-sig}^syn opcon + of unit
+%   ^\label{line:drop-sig}^syn opcon - of Lbl
+% }
+
+% ^\label{line:ltplx-tsm-start}^syntax ltpl(L :: LTUPLE, param :: ComponentSpec) at L.c(param) {
+%   static fn ps => (* 
+%     - elaborates {e_1, ..., e_n} to 
+%         L.intro_unlabeled () (e_1; ...; e_n)
+%     - elaborates {lbl1 => e_1, ..., lbln => e_n} to 
+%         L.intro_labeled [lbl1, ..., lbln] e_1 ... e_n
+%   *)
+% }^\label{ltplx-tsm-end}^
+
+% metamodule Ltuple :: LTUPLE = metamod {
+%   ^\label{line:c-start}^(* translate labeled tuple types to nested products *)
+%   tycon c {
+%   	(* type translation generator: *)
+%   	static fn (param :: ComponentSpec) :: ITy => fold (comp_spec_to_tuples param)
+%       `SCSSunitECSS`
+%       (fn ((lbl, ty), cur_ty_trans) => `SCSStrans(ECSStySCSS) * %ECSScur_ty_trans`)
+%   }^\label{line:c-end}^
+
+%   ^\label{line:introul-start}^ana opcon intro_unlabeled {
+%   	(* translate intro operator to nested pairs *)
+%   	static fn (ty :: Ty, op_param :: unit, args :: list(Arg)) :: IExp => tycase(ty) {
+%         c(ty_param) => (fold (zipExact (comp_spec_to_tuples ty_param) args)
+%       	  `SCSS()ECSS`
+%           (fn (((lbl, ty), arg), cur_trans) => `SCSS(%{ECSSana(arg, ty)SCSS}, %ECSScur_transSCSS)ECSS`)
+%         )
+%       | _ => raise TyErr("...")
+%     }
+%   }^\label{line:introul-end}^
+
+%   ^\label{line:introl-start}^syn opcon intro_labeled {
+%   	static fn (op_param :: list(Lbl), args :: list(Arg)) :: Ty * IExp => 
+%       (* ... similar to intro_unlabeled but reorder first ... *)
+%   }^\label{line:introl-end}^
+
+%   ^\label{line:prj-start}^syn opcon # {
+%   	static fn (op_param :: Lbl, args :: list(Arg)) :: Ty * IExp =>  
+%       (* ... generate the appropriate n-fold projection ... *)
+%   }^\label{line:prj-end}^
+
+%   ^\label{line:ltpl-conc-start}^syn opcon + {
+%   	static fn (op_param :: unit, args :: list(Arg)) :: Ty * IExp => 
+%       (* generate the new type and combine the two nested tuples *)
+%   }^\label{line:ltpl-conc-end}^
+
+%   ^\label{line:ltpl-drop-start}^syn opcon - {
+%   	static fn (op_param :: Lbl, args :: list(Arg)) :: Ty * IExp => 
+%       (* generate the new type and drop the appropriate component *)
+%   }^\label{line:ltpl-drop-end}^
+% } with syntax ltpl^\label{line:ltplx-annotation}^
+% (* synonyms used in Sec. 5.1 *)
+% ^\label{line:tycon-synonym}^let tycon ltuple = Ltuple.c
+% let opcon # = Ltuple.#
+% let opcon ltuple+ = Ltuple.+ 
+% let opcon ltuple- = Ltuple.-
+% \end{lstlisting}
+% \caption{Using metamodules to define the type structure of labeled tuple types.}
+% \label{fig:ltuplex}
+% \end{figure}

### 2015-10-09 thesis-proposal/problem.tex
proposal revisions

@@
 Instead, its users are able to modularly  express a wide variety of useful constructs in libraries, using the comparatively small set of primitives that the language builds in.
-Functional programming language designers have long studied small typed lambda calculi to develop a principled understanding of  fundamental metatheoretic issues, like type safety, and to examine the mathematical character of language primitives of interest in isolation.
-These studies have informed the design of ``full-scale''\footnote{Throughout this work, phrases that should be read as having an intuitive meaning, rather than a strict mathematical meaning, will be introduced with quotation marks.} functional languages, which combine several such primitives and also introduce various generalizations and primitive ``embellishments''  motivated by a consideration of human factors.
+Functional programming language designers have long studied small typed lambda calculi to develop a principled understanding of language-theoretic issues, like type safety, and to examine the mathematical character of language primitives of possible interest in isolation.
+These studies have then informed the design of ``full-scale''\footnote{Throughout this work, words and phrases that should be read as having an intuitive or informal meaning, rather than a strict mathematical meaning, will be introduced with quotation marks.} functional languages, which combine several such primitives and also introduce various generalizations and primitive ``embellishments''  motivated by a consideration of various human factors.
 For example, major functional languages like Standard ML (SML) \cite{mthm97-for-dart,harper1997programming}, OCaml \cite{ocaml-manual} and Haskell \cite{jones2003haskell} all primitively build in record types, generalizing the nullary and binary product types that suffice in simpler calculi, because explicitly labeled components are cognitively useful to human programmers.
@@
 one where programmers can direct their efforts almost exclusively toward the development of libraries for a wide variety of application domains.
-However, a stable language design that fully achieves this ideal has yet to emerge, as evidenced by the diverse array of ``dialects'' that continue to proliferate around all major contemporary languages.
+However, a stable language design that fully achieves this ideal has yet to emerge, as evidenced by the diverse array of ``dialect'' languages that continue to proliferate around all major contemporary languages.
 In fact, tools that aid in the construction of so-called  ``domain-specific'' language dialects (DSLs)\footnote{In some parts of the literature, such dialects are called ``external DSLs'', to distinguish them from  ``internal'' or ``embedded DSLs'', which are actually  library interfaces that only ``resemble'' distinct dialects \cite{fowler2010domain}.} seem only to be becoming increasingly prominent.
@@
 
-Although some of these dialects extend the semantics of the language that they are based on, the more common situation appears to be that the existing semantic primitives suffice and it is only the \emph{syntactic cost} of expressing one or more constructs of interest using these primitives  that is too high.
+Some of these dialects extend the type structure of the language that they are based on.
+However, the more common situation, and the one that we will be our focus in this work, appears to be that the existing semantic primitives suffice and it is only the \emph{syntactic cost} of expressing one or more constructs of interest using these primitives  that is too high.
 In response, library providers construct \emph{syntactic dialects} -- dialects that introduce only new derived syntactic forms.
@@
 Second, primitives introduced earlier in a language's lifespan can end up monopolizing finite ``syntactic resources'', forcing subsequent primitives to use ever more esoteric forms.
-And third, primitives that prove afteer some time to be flawed in some way cannot be removed or replaced without breaking backwards compatibility.
+And third, primitives that prove after some time to be flawed in some way cannot be removed or replaced without breaking backwards compatibility.
 %Because there is no data about how useful a construct is in practice until it is available in a major language, decisions about which constructs to include are often informed only by intuition.
@@
 \item \textbf{Type-specific languages} (TSLs), introduced in Sec.
-\ref{sec:tsls}, further reduce syntactic cost by allowing library providers to associate a TSM with a type declaration and then rely on a local type inference scheme to invoke it implicitly.
+\ref{sec:tsls}, further reduce syntactic cost by allowing library providers to associate a TSM with a type declaration and then rely on a local type inference scheme to invoke that TSM implicitly.
 %\item \textbf{Metamodules}, introduced in Sec.
@@
 
-The primitives we propose are forms of \emph{static code generation} (also sometimes called \emph{static} or \emph{compile-time metaprogramming}), meaning that the relevant rules in the static semantics of the language call for the evaluation of \emph{static functions} that generate static representations of terms.
-Static functions are functions written in a restricted subset of the language (we will discuss the necessary restrictions in Sec.
-\ref{sec:tsms-limitations}).%Library providers write these static functions using the Verse \emph{static language} (SL).
-
+The primitives we propose perform \emph{static code generation} (also sometimes called \emph{static} or \emph{compile-time metaprogramming}), meaning that the relevant rules in the static semantics of the language call for the evaluation of \emph{static functions} that generate static representations of terms.
+Static functions are functions written in a restricted subset of the language (we will discuss the design space of restrictions in Sec.
+\ref{sec:tsms-limitations}).
+%Library providers write these static functions using the Verse \emph{static language} (SL).
 The design we are proposing also has conceptual roots in earlier work on \emph{active libraries}, which similarly envisioned using compile-time computation to give library providers more control over aspects of the language (but did not take a type theoretic approach) \cite{active-libraries-thesis}.
@@
 
-\paragraph{Abstract Syntax} The abstract syntax of {patterns}, $r$, over strings, $s$, is specified as below:\[r ::= \textbf{empty} ~|~ \textbf{str}(s) ~|~ \textbf{seq}(r; r) ~|~ \textbf{or}(r; r) ~|~ \textbf{star}(r) ~|~ \textbf{group}(r)\]
+\paragraph{Abstract Syntax} The abstract syntax of {patterns}, $r$, over strings, $s$, is specified below:\[r ::= \textbf{empty} ~|~ \textbf{str}(s) ~|~ \textbf{seq}(r; r) ~|~ \textbf{or}(r; r) ~|~ \textbf{star}(r) ~|~ \textbf{group}(r)\]
 One way to express this abstract syntax is by defining a recursive sum type \cite{pfpl}. Verse supports these as datatypes:
@@
 It might be useful for patterns with the same normal form to be  indistinguishable from the perspective of client code.
-Second, it can be useful for performance reasons to maintain additional data alongside patterns (e.g.
+Second, it can be useful for performance reasons to maintain additional data alongside regexes (e.g.
 a corresponding finite automata) without exposing this ``implementation detail'' to clients.
@@
 \paragraph{Concrete Syntax} The abstract syntax of patterns is too verbose to be practical  in all but the most trivial examples, so programmers conventionally write patterns using a more concise concrete syntax.
-For example, the concrete syntax \lstinline{A|T|G|C} corresponds to the following much more verbose pattern expression (assuming some module \lstinline{R : RX} is in scope here and in the remainder of the examples):
+For example, the concrete syntax \lstinline{A|T|G|C} corresponds to the following much more verbose pattern expression (assuming some module \lstinline{R : RX} is in scope here and in the remainder of the document):
 \begin{lstlisting}[numbers=none,mathescape=|]

### 2015-10-11 thesis-proposal/problem.tex
removed various doubling

@@
 Functional programming language designers have long studied small typed lambda calculi to develop a principled understanding of language-theoretic issues, like type safety, and to examine the mathematical character of language primitives of possible interest in isolation.
-These studies have then informed the design of ``full-scale''\footnote{Throughout this work, words and phrases that should be read as having an intuitive or informal meaning, rather than a strict mathematical meaning, will be introduced with quotation marks.} functional languages, which combine several such primitives and also introduce various generalizations and primitive ``embellishments''  motivated by a consideration of various human factors.
+These studies have then informed the design of ``full-scale''\footnote{Throughout this work, words and phrases that should be read as having an intuitive or informal meaning, rather than a strict mathematical meaning, will be introduced with quotation marks.} functional languages, which combine several such primitives and also introduce various generalizations and primitive ``embellishments''  motivated by a consideration of human factors.
 For example, major functional languages like Standard ML (SML) \cite{mthm97-for-dart,harper1997programming}, OCaml \cite{ocaml-manual} and Haskell \cite{jones2003haskell} all primitively build in record types, generalizing the nullary and binary product types that suffice in simpler calculi, because explicitly labeled components are cognitively useful to human programmers.

### 2015-10-12 thesis-proposal/problem.tex
minor corrections

@@
 Instead, its users are able to modularly  express a wide variety of useful constructs in libraries, using the comparatively small set of primitives that the language builds in.
-Functional programming language designers have long studied small typed lambda calculi to develop a principled understanding of language-theoretic issues, like type safety, and to examine the mathematical character of language primitives of possible interest in isolation.
+Functional programming language designers have long studied small typed lambda calculi to develop a principled understanding of language-theoretic issues, like type safety, and to examine the mathematical character of language primitives of interest in isolation.
 These studies have then informed the design of ``full-scale''\footnote{Throughout this work, words and phrases that should be read as having an intuitive or informal meaning, rather than a strict mathematical meaning, will be introduced with quotation marks.} functional languages, which combine several such primitives and also introduce various generalizations and primitive ``embellishments''  motivated by a consideration of human factors.
@@
 
-Some of these dialects extend the type structure of the language that they are based on.
-However, the more common situation, and the one that we will be our focus in this work, appears to be that the existing semantic primitives suffice and it is only the \emph{syntactic cost} of expressing one or more constructs of interest using these primitives  that is too high.
+Although some of these dialects extend the type structure of the language that they are based on, the most common situation, and the one that will be our focus in this work, is that the existing semantic primitives suffice and it is only the \emph{syntactic cost} of expressing one or more constructs of interest using these primitives  that is too high.
 In response, library providers construct \emph{syntactic dialects} -- dialects that introduce only new derived syntactic forms.
@@
 }{
-	\Delta~\Gamma \vdash \lambda x{:}\tau.\hat{e} \leadsto \lambda x{:}\tau.e : \tau \rightharpoonup \tau'hh
+	\Delta~\Gamma \vdash \lambda x{:}\tau.\hat{e} \leadsto \lambda x{:}\tau.e : \tau \rightharpoonup \tau'
 }

### 2015-10-15 thesis-proposal/problem.tex
Updated link to thesis repo
fixed typo
fixed?

@@
 \section{Typed Syntax Macros (TSMs)}\label{sec:tsms}
-We will now introduce a new external primitive -- the \textbf{typed syntax macro} (TSM) -- that combines the syntactic flexibility of syntax extensions with the reasoning guarantees of typed macros.
+We will now introduce a new primitive -- the \textbf{typed syntax macro} (TSM) -- that combines the syntactic flexibility of syntax extensions with the reasoning guarantees of typed macros.
 Like the term-rewriting macros just described, TSMs can be parameterized by modules, so they can be used to define syntax valid at any abstract type defined by a module satisfying a specified signature.
@@
 
-\subsection{TSMs by Example}\label{sec:tsms-by-example}
+\subsection{TSMs By Example}\label{sec:tsms-by-example}
 %A typed syntax macro is invoked by applying it to a \emph{delimited form}, which can contain  arbitrary syntax in its \emph{body}.  
@@
 
-\url{https://github.com/cyrus-/papers/blob/master/thesis/omar-thesis.pdf}
+\url{https://github.com/cyrus-/thesis/blob/master/omar-thesis.pdf}
 %To validate our claims of expressiveness, I plan to give several more examples drawn from existing languages.

### 2015-11-02 thesis-proposal/problem.tex
Tightened text of proposal intro, contributions sections

@@
 Instead, its users are able to modularly  express a wide variety of useful constructs in libraries, using the comparatively small set of primitives that the language builds in.
-Functional programming language designers have long studied small typed lambda calculi to develop a principled understanding of language-theoretic issues, like type safety, and to examine the mathematical character of language primitives of interest in isolation.
-These studies have then informed the design of ``full-scale''\footnote{Throughout this work, words and phrases that should be read as having an intuitive or informal meaning, rather than a strict mathematical meaning, will be introduced with quotation marks.} functional languages, which combine several such primitives and also introduce various generalizations and primitive ``embellishments''  motivated by a consideration of human factors.
-For example, major functional languages like Standard ML (SML) \cite{mthm97-for-dart,harper1997programming}, OCaml \cite{ocaml-manual} and Haskell \cite{jones2003haskell} all primitively build in record types, generalizing the nullary and binary product types that suffice in simpler calculi, because explicitly labeled components are cognitively useful to human programmers.
-Similarly, these languages build in derived syntactic forms (colloquially, ``syntactic sugar'') that decrease the syntactic cost of working with common library constructs, like lists.
-
-The hope amongst many language designers is that a limited number of primitives like these will suffice to produce a ``general-purpose'' programming language, i.e.
-one where programmers can direct their efforts almost exclusively toward the development of libraries for a wide variety of application domains.
-However, a stable language design that fully achieves this ideal has yet to emerge, as evidenced by the diverse array of ``dialect'' languages that continue to proliferate around all major contemporary languages.
-In fact, tools that aid in the construction of so-called  ``domain-specific'' language dialects (DSLs)\footnote{In some parts of the literature, such dialects are called ``external DSLs'', to distinguish them from  ``internal'' or ``embedded DSLs'', which are actually  library interfaces that only ``resemble'' distinct dialects \cite{fowler2010domain}.} seem only to be becoming increasingly prominent.
-%\subsection{Why are there so many dialects?}
-%{This calls for an investigation}: why is it that programmers and researchers are still so often unable to satisfyingly express the constructs that they need in libraries, as modes of use of the ``general-purpose'' primitives already available in major languages today, and instead see a need for new language dialects?
-
-Although some of these dialects extend the type structure of the language that they are based on, the most common situation, and the one that will be our focus in this work, is that the existing semantic primitives suffice and it is only the \emph{syntactic cost} of expressing one or more constructs of interest using these primitives  that is too high.
-In response, library providers construct \emph{syntactic dialects} -- dialects that introduce only new derived syntactic forms.
-%Put another way, syntactic dialects can be specified by a context-independent expansion to the existing language that they are based on. 
-For example, Ur/Web is a syntactic dialect of Ur (a language that itself descends from ML \cite{conf/pldi/Chlipala10}) that builds in derived forms for SQL queries, HTML elements and other datatypes used in the domain of web programming \cite{conf/popl/Chlipala15}.
+Functional programming languages come in many sizes.
+The smallest functional languages -- often referred to as ``{lambda calculi}'' -- allow researchers and language designers to study the mathematical properties of language primitives of interest in isolation.
+These ``small-scale'' languages inform the design of ``full-scale''
+%\footnote{Throughout this work, words and phrases that should be read as having an intuitive or informal meaning, rather than a strict mathematical meaning, will be introduced with quotation marks.} 
+ languages, which combine several such primitives and also define various derived syntactic forms (colloquially, ``syntactic sugar'') that decrease the syntactic cost of selected idioms. 
+For example, Standard ML (SML) \cite{mthm97-for-dart,harper1997programming}, OCaml \cite{ocaml-manual} and Haskell \cite{jones2003haskell}   build in 
+%record types, generalizing the nullary and binary product types that suffice in simpler calculi, because labeled components are cognitively useful to human programmers.
+Similarly, these languages all build in
+derived forms that decrease the syntactic cost of working with lists. For example, in these languages, the form \lstinline{[1, 2, 3, 4, 5]} desugars to: 
+\begin{lstlisting}[numbers=none]
+Cons(1, Cons(2, Cons(3, Cons(4, Cons(5, Nil)))))
+\end{lstlisting}
+
+The hope amongst many language designers is that a limited number of derived forms like these will suffice to produce a ``general-purpose'' programming language, i.e.
+one that satisfies programmers working in a wide variety of application domains.
+Unfortunately, a stable language design that fully achieves this ideal has yet to emerge, as evidenced by the diverse array of \emph{syntactic dialects} -- dialects that introduce only new derived forms -- that continue to proliferate around all major contemporary languages.
+For example, Ur/Web is a syntactic dialect of Ur (an ML-like full-scale language \cite{conf/pldi/Chlipala10}) that builds in derived forms for SQL queries, HTML elements and other datatypes used in the domain of web programming \cite{conf/popl/Chlipala15}.
 %Syntactic cost is often assessed qualitatively \cite{green1996usability}, though quantitative metrics can be defined.
-This is not an isolated example -- there are many other types of data that could similarly benefit from the availability of specialized derived forms (we will consider another example, regular expression patterns, in Sec.
-\ref{sec:syntax-examples}).
-Tools like Camlp4 \cite{ocaml-manual}, Sugar* \cite{erdweg2011sugarj,erdweg2013framework} and Racket \cite{Flatt:2012:CLR:2063176.2063195}, which we will discuss in Sec.
-\ref{sec:syntax-existing-approaches}, have lowered the engineering costs of constructing syntactic dialects in such situations, further contributing to their proliferation.
-
-%More advanced dialects introduce new type structure, going beyond what is possible with only new derived forms.
-As a simple example, the static and dynamic semantics of records cannot be expressed by context-independent expansion to a language with only nullary and binary products.
-Various languages have explored ``record-like'' primitives that go further, supporting functional update operators, width and depth coercions (sometimes implicit)%\cite{Cardelli:1984:SMI:1096.1098}
-%, methods, prototypic dispatch and other such ``semantic embellishments'' that in turn cannot be expressed by context-independent expansion to a language with only standard record types (we will detail an  example in Sec.
-\ref{sec:metamodules-motivating-examples}).
-OCaml primitively builds in the type structure of polymorphic variants, open datatypes and  operations that use format strings like $\mathtt{sprintf}$ \cite{ocaml-manual}.
-ReactiveML builds in primitives for functional reactive programming \cite{mandel2005reactiveml}.
-ML5 builds in high-level primitives for distributed programming based on a modal lambda calculus \cite{Murphy:2007:TDP:1793574.1793585}.
-Manticore \cite{conf/popl/FluetRRSX07} and AliceML  \cite{AliceLookingGlass} build in parallel programming primitives with a more elaborate type structure than is found in simpler accounts of parallelism.
-%MLj builds in the type structure of the Java object system (motivated by a desire to interface safely and naturally with Java libraries) \cite{Benton:1999:IWW:317636.317791}.
-Other dialects do the same for other foreign languages, e.g.
-Furr and Foster describe a dialect of OCaml that builds in the type structure of C \cite{Furr:2005:CTS:1065010.1065019}.
-Tools like proof assistants and logical frameworks are used to specify and reason metatheoretically about dialects like these, and tools like compiler generators and language frameworks \cite{erdweg2013state} lower their implementation cost, again contributing to their proliferation.
+We will consider another example -- regular expression patterns -- in Sec. \ref{sec:examples}. 
+Tools like Camlp4 \cite{ocaml-manual}, Sugar* \cite{erdweg2011sugarj,erdweg2013framework} and Racket's preprocessor \cite{Flatt:2012:CLR:2063176.2063195}, which we will discuss in Sec.
+\ref{sec:syntax-existing-approaches}, have decreased the engineering costs of constructing syntactic dialects, further contributing to their proliferation.
 
@@
 % test 
-Some  view the ongoing proliferation of dialects described above as harmless or even as desirable, arguing that programmers can simply choose the right dialect for the job at hand \cite{journals/stp/Ward94}.
-However, this ``dialect-oriented'' approach is, in an important sense, anti-modular: a library written in one dialect cannot, in general, safely and idiomatically interface with a library written using another dialect.
-Addressing this interoperability problem requires somehow ``combining'' the  dialects into a single language \cite{Benton:1999:IWW:317636.317791}.
-However, in the most general setting where the dialects in question might be specified by judgements of arbitrary form, this is not a well-defined notion.
-Even if we restrict our interest to dialects specified using formalisms that do operationalize some notion of dialect combination, there is generally no guarantee that the combined dialect will conserve important metatheoretic properties that can be established about the dialects in isolation.
+Some  view this proliferation of dialects as harmless or even as desirable, arguing that programmers can simply choose the right dialect for the job at hand \cite{journals/stp/Ward94}.
+However, this ``dialect-oriented'' approach is, in an important sense, anti-modular: programmers cannot always ``combine'' different dialects if they want to use the primitives that they feature together in a single program.
+For example, a programmer might have access to a syntactic dialect featuring HTML syntax and one featuring regular expression syntax, but it may not be straightforward to construct a dialect featuring both.
+
+In some cases, coming up with a combined dialect is difficult because the dialects are specified using different formalisms.
+In other cases, a common formalism has been used, but it does not operationalize the notion of dialect combination (e.g.
+Racket's preprocessor \cite{Flatt:2012:CLR:2063176.2063195}).
+But even if we restrict our interest to dialects specified using a common formalism that does operationalize some notion of dialect combination, there may still be a problem: the formalism may not guarantee that the combined dialect will conserve important properties that can be established about the dialects in isolation.
 %In other words, any putative ``combined language'' must formally be considered a  distinct system for which one must derive essentially all metatheorems of interest anew, guided only informally by those derived for the dialects individually.
 %There is no well-defined mechanism for constructing such a ``combined language'' in general.
-For example, consider two syntactic dialects, one specifying derived syntax for finite mappings, the other specifying a similar syntax for \emph{ordered} finite mappings.
-Though each dialect can be specified by an unambiguous grammar in isolation, when these grammars are na\"ively  combined by, for example, Camlp4,  ambiguities  arise.
+For example, consider two syntactic dialects specified using Camlp4, one specifying derived syntax for finite mappings, the other specifying overlapping syntax for \emph{ordered} finite mappings.
+Though each dialect has an unambiguous (a.k.a.
+deterministic) grammar, when these grammars are na\"ively  combined,  ambiguities will arise.
+We are aware of only one formalism that guarantees that determinism is conserved when syntactic dialects are combined \cite{conf/pldi/SchwerdfegerW09}, but it has limited expressive power, as we will discuss in Sec.
+\ref{sec:direct-syntax-extension}.
 %It is thus infeasible to simply allow different contributors to a software system to choose their own favorite dialect for each component they are responsible for. 
 %It it clear that dialects are better rhetorical devices than practical engineering artifacts.
-Due to this paucity of modular reasoning principles, the ``dialect-oriented'' approach is problematic for software development ``in the large''.
+
+%Due to this paucity of modular reasoning principles, the ``dialect-oriented'' approach is problematic for software development ``in the large''.
 %Large software projects and software ecosystems must pick a single language that does provide powerful modular reasoning principles and, to benefit from them, stay inside it.
 
-Dialect designers have instead had to take a less direct approach to have an impact on large-scale software development: they have had to convince the designers in control of comparatively popular languages, like OCaml and Scala, to include some suitable variant of the primitives they've espoused into backwards compatible language revisions.
+\subsection{Large Languages Considered Harmful}
+Dialects sometimes have a less direct influence on large-scale software development: they can help convince the designers in control of comparatively popular languages, like OCaml and Scala, to include some variant of the primitives that they feature into backwards-compatible language revisions.
 %These decisions are increasingly influenced by community processes, e.g.
@@
 This \emph{ad hoc} approach is not sustainable, for three main reasons.
-First,  there are simply too  many potentially useful such primitives, and many of these are only relevant in relatively narrow application domains (for derived syntax, our group has  gathered initial data speaking to this \cite{TSLs}).
+First, we have gathered empirical data that suggests that there are simply too  many potentially useful such primitives \cite{TSLs}.
+Many of these are only relevant in relatively narrow application domains.
+It is unreasonable to expect language designers to be able to evaluate all of these use cases in a timely and informed manner.
 Second, primitives introduced earlier in a language's lifespan can end up monopolizing finite ``syntactic resources'', forcing subsequent primitives to use ever more esoteric forms.
-And third, primitives that prove after some time to be flawed in some way cannot be removed or replaced without breaking backwards compatibility.
-%Because there is no data about how useful a construct is in practice until it is available in a major language, decisions about which constructs to include are often informed only by intuition.
+And third, primitives that prove after some time to be flawed in some way cannot be removed or modified without breaking backwards compatibility.
+For all these reasons, language designers are justifiably reticent to add new primitives to major programming languages.%Because there is often no empirical data about how useful a construct is in practice until it is available in a major language, decisions about which constructs to include are often informed only by intuition (and are thus)
 %Recalling the words of  Reynolds, which are clearly as relevant today as they were almost half a century ago \cite{Reynolds70}: %This approach is antithetical to the ideal of a truly \emph{general-purpose language} described at the beginning of this section.
 %\newpage
-%\begin{quote}\textit{The recent development of programming languages suggests that the simul\-taneous achievement of simplicity 
-%and generality in language design is a serious unsolved 
-%problem.}\begin{flushright}--- John Reynolds (1970)\end{flushright}
-%\end{quote}
-
-This suggests that language designers should strive to keep general-purpose languages small, stable and free of \emph{ad hoc} primitives.
-This leaves  two possible paths forward.
-One, exemplified (arguably) by SML, is to simply eschew further ``embellishments'' and settle on the existing primitives, which might be considered to sit at a ``sweet spot'' in the overall language design space.
-% (accepting that in some circumstances, this trades away expressive power or leads to  high syntactic cost).
-The other path forward is to search for a small number of highly general primitives that allow us degrade many of the constructs that are built primitively into dialects today instead to modularly composable library constructs.
+
+\subsection{Toward More General Primitives}
+If, as we've argued, language designers should strive both to avoid dialect formation and to keep general-purpose languages small, stable and free of \emph{ad hoc}  primitives, this leaves two possible paths forward.
+One, exemplified (arguably) by SML, is to simply eschew ``niche'' embellishments and settle on the existing design, which might be considered to sit at a ``sweet spot'' in the overall language design space (accepting that in some circumstances, this leads to  high syntactic cost).
+The other path forward is to search for a small number of highly general primitives that allow us degrade many of the constructs that are built primitively into languages and their dialects today instead to modular library constructs.
 Encouragingly, primitives of this sort do occasionally arise.
 For example, a recent revision of OCaml added support for  generalized algebraic data types (GADTs), based on research on guarded recursive datatype constructors \cite{XiCheChe03}.
-Using GADTs, OCaml was able to move some of the \emph{ad hoc} machinery for typechecking operations that use format strings, like \texttt{sprintf}, out of the language and into a library.
-However, syntactic machinery remains  built in.
-
-%Similarly, it recently introduced ``open datatypes'', which subsume its previous more specialized exception type, and captures many use cases for .
-
-%Viewed ``dually'', one might equivalently ask for a language that builds in a core that is as small as possible, but provides expressive power comparable to languages with much larger cores.
-This is our goal in the work being proposed.
-
-%\vspace{-10px}
+Using GADTs, OCaml was able to move some of the \emph{ad hoc} machinery for typechecking operations that use format strings, like \li{sprintf}, out of the language and into a library.
+Syntactic machinery related to \li{sprintf}, however, remains  built in.
+
 \section{Proposed Contributions}
@@
 In particular, we plan to introduce the following primitives:
-% By supporting the primitives that we introduce, 1) Verse will be smaller than comparable languages like ML and Scala, and 2) dialect formation will be less frequently necessary.
+% By supporting the primitives that we introduce, 1) VerseML will be smaller than comparable languages like ML and Scala, and 2) dialect formation will be less frequently necessary.
 In other words, these primitives reduce the need for many others:%This eliminates the needs to build in fewer \emph{ad hoc} constructs and dialects are less frequently necessary.
@@
 
-%Verse features a module system taken directly from SML.
-Unlike SML, the Verse core language is split into a \emph{typed external language} (EL) specified by {type-directed translation} to a minimal \emph{typed internal language} (IL).
+%VerseML features a module system taken directly from SML.
+Unlike SML, the VerseML core language is split into a \emph{typed external language} (EL) specified by {type-directed translation} to a minimal \emph{typed internal language} (IL).
 \begin{enumerate}
@@
 \ref{sec:metamodules}, reduce the need to primitively build in the type structure of constructs like records (and variants thereof),  labeled sums and other interesting constructs that we will introduce later by giving library providers programmatic ``hooks'' directly into the semantics, which are specified as a \emph{type-directed translation semantics} targeting a small \emph{typed internal language} (introduced in Sec.
-\ref{sec:verse}).
+\ref{sec:VerseML}).
 %For example, a library provider can implement the type structure of records with a metamodule that:
@@
 As vehicles for this work, we plan to formally specify a series of small typed lambda calculi that capture each of the novel primitives that we introduce ``minimally''.
-For the sake of examples, we will also describe (but not formally specify) a ``full-scale'' functional language called Verse.\footnote{We distinguish Verse from Wyvern, which is the language referred to in prior publications about some of the work that we will describe, because Wyvern is a group effort evolving independently in some important ways.} Verse is based semantically on Standard ML, differing mainly in that it uses a local type inference scheme \cite{Pierce:2000:LTI:345099.345100} (like, for example, Scala) for reasons that we will return to in Sec.
+For the sake of examples, we will also describe (but not formally specify) a ``full-scale'' functional language called VerseML.\footnote{We distinguish VerseML from Wyvern, which is the language referred to in prior publications about some of the work that we will describe, because Wyvern is a group effort evolving independently in some important ways.} VerseML is, as its name suggests, a dialect of ML.
+It diverges from other dialects of ML that have a similar underlying type structure, like Standard ML and OCaml, in that it uses a local type inference scheme \cite{Pierce:2000:LTI:345099.345100} (like, for example, Scala \cite{OdeZenZen01}) for reasons that have to do with the mechanisms described in Sec.
 \ref{sec:tsls}.
-The reason we will not follow Standard ML \cite{mthm97-for-dart} in giving a complete formal specification of Verse is both to emphasize that the primitives we introduce can be considered for inclusion in a variety of language designs, and to avoid distracting the reader with specifications for ``orthogonal'' primitives that are already well-understood in the literature.
+The reason we will not follow Standard ML \cite{mthm97-for-dart} in giving a complete formal specification of VerseML in this work is both to emphasize that the primitives we introduce can be considered for inclusion in a variety of language designs (not exclusively dialects of ML), and to avoid distracting the reader with specifications for ``orthogonal'' primitives that are already well-understood in the literature.
 %We anticipate that future full-scale language specifications will be able to combine the ideas  in the proposed work without trouble.
 %The purpose of the work being proposed is to serve as a reference for those interested in the new constructs we introduce, not to serve as a language specification.
-%We will give a brief overview of these languages are organized in Sec. \ref{sec:verse}.
+%We will give a brief overview of these languages are organized in Sec. \ref{sec:VerseML}.
 
@@
 \ref{sec:tsms-limitations}).
-%Library providers write these static functions using the Verse \emph{static language} (SL).
+%Library providers write these static functions using the VerseML \emph{static language} (SL).
 The design we are proposing also has conceptual roots in earlier work on \emph{active libraries}, which similarly envisioned using compile-time computation to give library providers more control over aspects of the language (but did not take a type theoretic approach) \cite{active-libraries-thesis}.
-%Maintaining a separation between the static (or ``compile-time'') phase and the dynamic (or ``run-time'') phase is an important facet of Verse's design.
+%Maintaining a separation between the static (or ``compile-time'') phase and the dynamic (or ``run-time'') phase is an important facet of VerseML's design.
 % static code generation.
-%We will  also introduce a simple variant of each of these primitives that leverages Verse's support for local type inference to further reduce syntactic cost in certain common situations.
-
-The main challenge in the design of these primitives will come in ensuring that they are metatheoretically well-behaved.
-If we are not careful, many of the problems  that arise when combining language dialects, discussed earlier, could simply shift into the semantics of these primitives.\footnote{This is why languages  like Verse are often called ``extensible languages'', though this is somewhat of a misnomer.
+%We will  also introduce a simple variant of each of these primitives that leverages VerseML's support for local type inference to further reduce syntactic cost in certain common situations.
+
+The main challenge in the design of the primitives we introduce will come in ensuring that they are metatheoretically well-behaved.
+If we are not careful, many of the problems  that arise when combining language dialects, discussed earlier, could simply shift into the semantics of these primitives.\footnote{This is why languages  like VerseML are often called ``extensible languages'', though this is somewhat of a misnomer.
 The defining characteristic of an extensible language is that it \emph{doesn't} need to be extended in situations where other languages would need to be extended.
-We will avoid this somewhat confusing terminology.} Our main technical contributions will be in rigorously showing how to address these problems in a principled manner.
-In particular, syntactic conflicts will be impossible by construction and the semantics will validate code statically generated by TSMs and TSLs to maintain:
+We will avoid this somewhat confusing terminology.} Our main technical contributions will be to rigorously address these problems in a principled manner.
+In particular, we will maintain:
 \begin{itemize}
-\item a \emph{hygienic type discipline}, meaning that the language is type safe, that one can reason about the type of a well-typed expression without examining its expansion, and that the expansion does not make any assumptions about the names of variables in the surrounding context; and
+\item a \emph{type discipline}, meaning that the language is type safe, and that programmers reading a well-typed expression can determine its type  without examining its expansion; 
+\item a \emph{hygiene discipline} (a.k.a.
+a \emph{binding discipline}), meaning that the expansion logic does not make any assumptions about the names of variables in the context surrounding the expansion, nor does it introduce ``hidden bindings'' into subexpressions; and
 \item \emph{modular reasoning principles}, meaning that library providers will have the ability to reason about the constructs that they have defined in isolation, and clients will be able to use them safely in any combination, without the possibility of conflict.\footnote{This is not quite true -- simple naming conflicts can arise.
@@
 A functional programming language can give library providers the ability to %meta\-pro\-gram\-matic\-ally 
-express new syntactic expansions while maintaining a hygienic type discipline and modular reasoning principles.
-%These  primitives are  expressive enough to subsume the need for a variety of primitives that are, or would need to be, built in to comparable contemporary languages.
+express new syntactic expansions while maintaining a type discipline, a hygiene discipline and modular reasoning principles.%These  primitives are  expressive enough to subsume the need for a variety of primitives that are, or would need to be, built in to comparable contemporary languages.
 \end{quote}
@@
 We aim only to decrease the need for syntactic dialects in this work.
-% out a larger design space within a single language, Verse.%a subset of constructs that can be specified by a semantics of a certain ``shape'' specified by Verse (we will make this more specific later).
-%There is nothing ``universal'' about Verse.
+% out a larger design space within a single language, VerseML.%a subset of constructs that can be specified by a semantics of a certain ``shape'' specified by VerseML (we will make this more specific later).
+%There is nothing ``universal'' about VerseML.
 
 It may also be useful to explicitly acknowledge that library providers could leverage the primitives we introduce   to define constructs that are in ``poor taste''.
-We  expect that in practice, Verse will come with a standard library defining a carefully curated collection of standard constructs, as well as guidelines for advanced users regarding when it would be sensible to use the mechanisms we introduce (following the example of languages that support operator overloading or type classes \cite{Hall:1996:TCH:227699.227700}, which also have the potential for ``abuse'').
-For most programmers, using Verse should not be substantially different from using a language like ML or one of its dialects.%The vast majority of programmers should not use the primitives that we introduce directly.
-
-Finally, Verse intentionally is not a dependently-typed language like Coq, Agda or Idris, because these languages do not maintain a phase separation between ``compile-time'' and ``run-time.'' This phase separation is useful for programming tasks (where one would like to be able to discover errors before running a program, particularly programs that may have an effect) but less so for theorem proving tasks (where it is mainly the fact that a pure expression is well-typed that is of interest, by the propositions-as-types principle).
-Verse is designed to be used for programming tasks where SML, OCaml, Haskell or Scala would be used today, not for advanced theorem proving tasks.
+We  expect that in practice, VerseML will come with a standard library defining a carefully curated collection of standard constructs, as well as guidelines for advanced users regarding when it would be sensible to use the mechanisms we introduce (following the example of languages that support operator overloading or type classes \cite{Hall:1996:TCH:227699.227700}, which also have the potential for ``abuse'').
+For most programmers, using VerseML should not be substantially different from using a language like ML or one of its dialects.%The vast majority of programmers should not use the primitives that we introduce directly.
+
+Finally, VerseML intentionally is not a dependently-typed language like Coq, Agda or Idris, because these languages do not maintain a phase separation between ``compile-time'' and ``run-time.'' This phase separation is useful for programming tasks (where one would like to be able to discover errors before running a program, particularly programs that may have an effect) but less so for theorem proving tasks (where it is mainly the fact that a pure expression is well-typed that is of interest, by the propositions-as-types principle).
+VerseML is designed to be used for programming tasks where SML, OCaml, Haskell or Scala would be used today, not for advanced theorem proving tasks.
 That said, we conjecture that the primitives we describe could be added to languages like Gallina (the ``external language'' of the Coq proof assistant  \cite{Coq:manual}) with  modifications, but do not plan to pursue this line of research in this dissertation.
@@
 
-% \section{Language Overview}\label{sec:verse}
-% Let us begin with a brief overview of how Verse and the minimal calculi that we will develop are organized and specified at a high level.
-We will assume that readers are familiar with the early chapters of a textbook like \emph{TAPL} \cite{tapl} or \emph{PFPL} \cite{pfpl}.%Verse consists of a \emph{module language} atop a \emph{core language}.
+% \section{Language Overview}\label{sec:VerseML}
+% Let us begin with a brief overview of how VerseML and the minimal calculi that we will develop are organized and specified at a high level.
+We will assume that readers are familiar with the early chapters of a textbook like \emph{TAPL} \cite{tapl} or \emph{PFPL} \cite{pfpl}.%VerseML consists of a \emph{module language} atop a \emph{core language}.
 
 % \subsection{Core Language}
-% The Verse core language -- the language of types and expressions -- will be the focus of our efforts.
+% The VerseML core language -- the language of types and expressions -- will be the focus of our efforts.
 Figure \ref{fig:overview} gives a diagrammatic overview of how the core language is organized.
@@
 we make a judgemental distinction between \emph{type synthesis} (the type is an ``output'') and \emph{type analysis} (the type is an ``input'') \cite{Pierce:2000:LTI:345099.345100}.
-This will  allow us to explicitly specify how Verse's \emph{local type inference} works, and in particular, how it interacts with the primitives that we aim to introduce.
+This will  allow us to explicitly specify how VerseML's \emph{local type inference} works, and in particular, how it interacts with the primitives that we aim to introduce.
 Like Scala, we do not seek to support non-local type inference.
@@
 
-% %The main novelty of our work is that the Verse EL builds in very little syntax or type structure \emph{a priori}.
+% %The main novelty of our work is that the VerseML EL builds in very little syntax or type structure \emph{a priori}.
 %Instead, \textbf{metamodules} and \textbf{typed syntax macros} (TSMs) are expressive enough to render this unnecessary.
@@
 
-% The Verse module system is taken directly (up to small syntactic changes) from Standard ML, with which we assume a working familiarity for the purposes of this proposal \cite{harper1997programming,MacQueen:1984:MSM:800055.802036} (a related module system, e.g.
+% The VerseML module system is taken directly (up to small syntactic changes) from Standard ML, with which we assume a working familiarity for the purposes of this proposal \cite{harper1997programming,MacQueen:1984:MSM:800055.802036} (a related module system, e.g.
 OCaml's, would work just as well for our purposes).
@@
 %To begin, let us assume a standard, ML-like semantics for both the EL and SL and consider the EL's concrete syntax. 
-Verse, like most contemporary languages, specifies a textual concrete syntax.\footnote{Although Wyvern specified a layout-sensitive concrete syntax, to avoid unnecessary distractions, we will describe a more conventional layout-insensitive concrete syntax for Verse.} %We have chosen to specify a layout-sensitive textual concrete syntax (i.e.
+VerseML, like most contemporary languages, specifies a textual concrete syntax.\footnote{Although Wyvern specified a layout-sensitive concrete syntax, to avoid unnecessary distractions, we will describe a more conventional layout-insensitive concrete syntax for VerseML.} %We have chosen to specify a layout-sensitive textual concrete syntax (i.e.
 newlines and indentation are not ignored).
 This design choice is not  fundamental to our proposed contributions, but it will be useful for cleanly expressing a class of examples that we plan to discuss later.
-We plan to specify some novel aspects of Verse's concrete syntax with an \emph{Adams grammar} \cite{Adams:2013:PPI:2429069.2429129} (such a specification for Wyvern, which has a very similar syntax, can be found in \cite{TSLs}), but for the purposes of this proposal, we will simply introduce Verse's concrete syntax by example as we go on.
+We plan to specify some novel aspects of VerseML's concrete syntax with an \emph{Adams grammar} \cite{Adams:2013:PPI:2429069.2429129} (such a specification for Wyvern, which has a very similar syntax, can be found in \cite{TSLs}), but for the purposes of this proposal, we will simply introduce VerseML's concrete syntax by example as we go on.
 %For constructs that have an obvious analog in ML, we will omit a detailed explanation.
@@
 
-Verse takes a less \emph{ad hoc} approach -- rather than privileging  particular library constructs with primitive syntactic support, Verse exposes primitives that allow library providers to introduce new expansion logic on their own, in a safe and modular manner.
+VerseML takes a less \emph{ad hoc} approach -- rather than privileging  particular library constructs with primitive syntactic support, VerseML exposes primitives that allow library providers to introduce new expansion logic on their own, in a safe and modular manner.
 %Lists need no special consideration from the language specification.
-%The purpose of this section is to  a and then introduce Verse's syntax extension mechanisms.
+%The purpose of this section is to  a and then introduce VerseML's syntax extension mechanisms.
 %For forms with a clear analogy to a form in Standard ML, we will assume the  semantics are analagous without providing details.
@@
 In Secs.
-\ref{sec:tsms} and \ref{sec:tsls}, we introduce our proposed alternatives -- \emph{typed syntax macros} (TSMs) and the related \emph{type-specific languages} (TSLs) -- and discuss how they resolve these issues (as well as some limitations that they have).
+\ref{sec:tsms} and \ref{sec:tsls}, we introduce our proposed alternatives -- \emph{typed syntax macros} (TSMs) and the related \emph{type-specific languages} (TSLs) -- and discuss how they resolve these issues.
 We  also give an overview of how TSMs are formally specified.
-We give a concrete timeline for the remaining work in Sec.
+We provide a concrete timeline for the remaining work in Sec.
 \ref{sec:syntax-timeline}, and conclude in Sec.
@@
 \paragraph{Abstract Syntax} The abstract syntax of {patterns}, $r$, over strings, $s$, is specified below:\[r ::= \textbf{empty} ~|~ \textbf{str}(s) ~|~ \textbf{seq}(r; r) ~|~ \textbf{or}(r; r) ~|~ \textbf{star}(r) ~|~ \textbf{group}(r)\]
-One way to express this abstract syntax is by defining a recursive sum type \cite{pfpl}. Verse supports these as datatypes:
+One way to express this abstract syntax is by defining a recursive sum type \cite{pfpl}. VerseML supports these as datatypes:
 
@@
 Indeed, there may be many ways to represent regular expression patterns, each with different performance trade-offs.
-For these reasons, a better approach in Verse, as in ML, is to abstract over the choice of representation using  the module system's support for type abstraction.
+For these reasons, a better approach in VerseML, as in ML, is to abstract over the choice of representation using  the module system's support for type abstraction.
 In particular, we can define the following \emph{module signature}, where the type of patterns, \lstinline{t}, is held abstract:
@@
 % \paragraph{Patterns as Recursive Sums} A more semantically justifiable strategy is to define \lstinline{Pattern} as a recursive sum type  \cite{pfpl}.
-Verse supports these in the form of case types, which are analagous to datatypes in ML:
+VerseML supports these in the form of case types, which are analagous to datatypes in ML:
 
@@
 
-% The dialect of Standard ML implemented by SML/NJ, recognizing that directly using constructors is often less than ideal, provides a facility for quotation/antiquotation \cite{SML/Quote}, which if introduced into Verse could be used as follows:
+% The dialect of Standard ML implemented by SML/NJ, recognizing that directly using constructors is often less than ideal, provides a facility for quotation/antiquotation \cite{SML/Quote}, which if introduced into VerseML could be used as follows:
 % \begin{lstlisting}[numbers=none,escapechar=|,mathescape=|]
@@
 
-\subsubsection{Direct Syntax Extension}\label{sec:syntax-extension}
+\subsubsection{Direct Syntax Extension}\label{sec:syntax-extension}\label{sec:direct-syntax-extension}
 One tempting alternative to dynamic string parsing is to use a system that gives the users of a language the power to directly extend its concrete syntax with new derived forms.
@@
 the common URI convention used on the web and by the Java packaging system.
-However, this forces us to use a more verbose token like \lstinline{edu_cmu_verse_rx}.
+However, this forces us to use a more verbose token like \lstinline{edu_cmu_VerseML_rx}.
 There is no simple way for clients of our extension to define scoped abbreviations for starting tokens because this mechanism operates purely at the level of the context-free grammar.
@@
 \begin{lstlisting}[numbers=none]
-let val ssn = edu_cmu_verse_rx R /SURL\d\d\d-\d\d-\d\d\d\dEURL/
+let val ssn = edu_cmu_VerseML_rx R /SURL\d\d\d-\d\d-\d\d\d\dEURL/
 \end{lstlisting}
@@
 %A typed syntax macro is invoked by applying it to a \emph{delimited form}, which can contain  arbitrary syntax in its \emph{body}.  
-To introduce TSMs, consider the following concrete external expression:
+To introduce TSMs, consider the following concrete expression:
 \begin{lstlisting}[numbers=none]
@@
 Here, we apply a \emph{parameterized TSM}, \lstinline{rx}, first to a module parameter, \lstinline{R}, then to a \emph{delimited form}, \lstinline{/SURLA|T|G|CEURL/}.
-A number of alternative delimiters are also available in Verse's concrete syntax and could equivalently have been used, e.g.
+A number of alternative delimiters are also available in VerseML's concrete syntax and could equivalently have been used, e.g.
 \lstinline{<SURLA|T|G|CEURL>}.
@@
 another expression.
-In this case, \lstinline{pattern} generates the following expansion (written concretely):
+In this case, \lstinline{rx} generates the following expansion (written concretely):
 
@@
 The parse function is a static function of type \lstinline{Body -> Exp}.
-Both of these types are defined in the Verse \emph{prelude}, which is a set of definitions available ambiently.
+Both of these types are defined in the VerseML \emph{prelude}, which is a set of definitions available ambiently.
 The type \lstinline{Body} gives the static function access to the {body} of the delimited form (e.g.
@@
 \inferrule[T-U-abs]{
-	\Delta \vdash \tau_1~\mathtt{type}\\
+	\Delta \vdash \tau~\mathtt{type}\\
 	\Delta~\Gamma, x:\tau \vdash \hat{e} \leadsto e : \tau'
@@
 	\Delta \vdash \tau~\mathtt{type}\\
-	\emptyset~\emptyset \vdash \hat{e}_\text{parse} \leadsto e_\text{parse} : \mathsf{Body} \rightharpoonup \mathsf{Exp}
+	\emptyset~\emptyset \vdash \hat{e}_\text{parse} \leadsto e_\text{parse} : \mathsf{Body} \rightharpoonup \mathsf{ExpOption}
 }{
@@
 	\Delta \vdash \tau~\mathtt{type}\\
-	\emptyset~\emptyset \vdash \hat{e}_\text{parse} \leadsto e_\text{parse} : \mathsf{Body} \rightharpoonup \mathsf{Exp}\\\\
+	\emptyset~\emptyset \vdash \hat{e}_\text{parse} \leadsto e_\text{parse} : \mathsf{Body} \rightharpoonup \mathsf{ExpOption}\\\\
 	b \downarrow e_\text{body} : \mathsf{Body}\\
-	e_\text{parse}(e_\text{body}) \Downarrow e_\text{exp}\\
+	e_\text{parse}(e_\text{body}) \Downarrow \mathsf{Some}(e_\text{exp})\\
 	e_\text{exp} : \mathsf{Exp} \uparrow \dot{e}_\text{exp}\\\\
@@
 \ref{sec:examples}.
-We can pattern match over a value, \lstinline{r}, of type \lstinline{Rx} using Verse's \lstinline{match} construct like this:
+We can pattern match over a value, \lstinline{r}, of type \lstinline{Rx} using VerseML's \lstinline{match} construct like this:
 \begin{lstlisting}[numbers=none]
@@
 To further lower the syntactic cost of using TSMs, so that it compares to the syntactic cost of derived syntax built in primitively (e.g.
-list syntax), we will now discuss how Verse allows library providers to define \emph{type-specific languages} (TSLs) by associating a TSM directly with an abstract type or datatype.
+list syntax), we will now discuss how VerseML allows library providers to define \emph{type-specific languages} (TSLs) by associating a TSM directly with an abstract type or datatype.
 When the type system encounters a delimited form not prefixed by a TSM name, it applies the TSM associated with the type it is being analyzed against implicitly.
@@
 %To validate our claims of expressiveness, I plan to give several more examples drawn from existing languages.
-I will include those we have worked out in the papers just cited as well as additional examples that highlight the benefits of integration with Verse's ML-based module system.
+I will include those we have worked out in the papers just cited as well as additional examples that highlight the benefits of integration with VerseML's ML-based module system.
 %For example, I will discuss implementing Haskell's primitive ``do'' syntax using TSLs together with a modular encoding of monads (as described on Prof.
@@
 % The function in the indented block is invoked {statically} to generate an \emph{elaboration} based on the {body} of the \emph{delimited form} provided to the TSM, here \lstinline{SHTML<h1><%EHTMLxSHTML></h1>EHTML},
-% for subsequent typechecking and translation to the Verse IL.
+% for subsequent typechecking and translation to the VerseML IL.
 In this example, the elaboration generated is equivalent to the one shown above in \lstinline{make_heading}.
-Both \lstinline{ParseStream}, which provides access to the body as a stream of characters, and \lstinline{Exp}, which encodes the abstract syntax of the Verse EL, are types defined in the Verse \emph{prelude}, which is simply a collection of definitions loaded before all others.
+Both \lstinline{ParseStream}, which provides access to the body as a stream of characters, and \lstinline{Exp}, which encodes the abstract syntax of the VerseML EL, are types defined in the VerseML \emph{prelude}, which is simply a collection of definitions loaded before all others.
 The implementation of this parse function is not particularly interesting, so we omit it here.
@@
 The TSM determines what syntactically distinguishes spliced expressions (here, the delimiters \lstinline{<%} and \lstinline{>} around an identifier or, not shown, \lstinline{<%>} and \lstinline{</%>} around an expression).
-% Verse only specifies the delimiters that can be used to transition from the expression language to the TSM body (here, we used backticks, but Verse supports several others, including \emph{layout delimiters}, which we will discuss later).
+% VerseML only specifies the delimiters that can be used to transition from the expression language to the TSM body (here, we used backticks, but VerseML supports several others, including \emph{layout delimiters}, which we will discuss later).
 This ensures that syntactic conflicts cannot arise by construction and localizes reasoning about syntactic ambiguity to each TSM, without incurring substantial syntactic cost.
@@
 Not shown here, this annotation can be parameterized by types as well as modules.
-Because the Verse module system is borrowed directly from ML,  this means that TSMs can be used to define derived syntax valid for all instances of an abstract type.
+Because the VerseML module system is borrowed directly from ML,  this means that TSMs can be used to define derived syntax valid for all instances of an abstract type.
 We will discuss this after motivating it with another example in Section \ref{sec:syntax}.
 
-% To decrease the syntactic cost of applying a TSM even further (controlling syntactic cost being, of course, the very purpose of TSMs), Verse can often locally infer these parameters.
-Moreover, when a type is declared, Verse allows the library provider to designate one TSM as its \emph{type-specific language} (TSL).
+% To decrease the syntactic cost of applying a TSM even further (controlling syntactic cost being, of course, the very purpose of TSMs), VerseML can often locally infer these parameters.
+Moreover, when a type is declared, VerseML allows the library provider to designate one TSM as its \emph{type-specific language} (TSL).
 Clients can then omit mention of this TSM entirely in positions where an expression of that type is expected.
@@
 
-%These mechanisms are one reason why we specify Verse by a bidirectionally typed elaboration semantics.
+%These mechanisms are one reason why we specify VerseML by a bidirectionally typed elaboration semantics.
 
@@
 % In the previous section, we assumed that the EL had an otherwise standard, ML-like type structure and considered only TSMs and TSLs as departures from ML.
-In fact, constructs like datatypes (shown briefly in use in the previous section), tuples, records and objects are not built primitively into Verse.
+In fact, constructs like datatypes (shown briefly in use in the previous section), tuples, records and objects are not built primitively into VerseML.
 Instead, they can be expressed using \emph{metamodules} (which are distinct from, though closely related to, {modules}, as we will discuss below).
-Only  functions are built primitively into the Verse EL.
+Only  functions are built primitively into the VerseML EL.
 
@@
 % \subsection{Existing Approaches}\label{sec:metamodules-related}
-% Before describing how we express the type structure of labeled tuples and regular strings in Verse using metamodules, let us consider some alternative approaches for each.
+% Before describing how we express the type structure of labeled tuples and regular strings in VerseML using metamodules, let us consider some alternative approaches for each.
 
 % \subsubsection{Labeled Tuples}
-% Recall that Verse builds only nullary and binary products into the IL, so we cannot express the type structure described above for labeled tuples  directly using IL primitives.
-The Verse EL does not build in even these, but so as to separate concerns, let us assume for the moment that regular string types are available in the EL: %So to consider existing approaches, we have to at least somewhat weaken our insistence on minimalism.
+% Recall that VerseML builds only nullary and binary products into the IL, so we cannot express the type structure described above for labeled tuples  directly using IL primitives.
+The VerseML EL does not build in even these, but so as to separate concerns, let us assume for the moment that regular string types are available in the EL: %So to consider existing approaches, we have to at least somewhat weaken our insistence on minimalism.
 % \begin{lstlisting}[numbers=none]
@@
 
-% Even if we were to automate this (using, for example, \emph{type macros} as  in Scala \cite{ScalaMacros2013}, adapted to support Verse's modules), another issue is that we can only reasonably express the introductory and projection operators by specifying their semantics as functions like this.
+% Even if we were to automate this (using, for example, \emph{type macros} as  in Scala \cite{ScalaMacros2013}, adapted to support VerseML's modules), another issue is that we can only reasonably express the introductory and projection operators by specifying their semantics as functions like this.
 To enumerate all possible operators that arise from the opcon \lstinline{ltuple-} using functions would require constructing an encoding of every labeled tuple type that might arise from dropping one or more components from the labeled tuple in question.
@@
 % If, instead of this module-based encoding, we further weaken our insistence on minimalism and primitively build record and tuple types into the EL, as SML and many other languages have done, let us consider what might be possible.
-Note that these too are library constructs in Verse, but because they are so commonly built in to other languages, this situation is worth considering.%we can retreat from trying to use the module system, but have only slightly better options.
+Note that these too are library constructs in VerseML, but because they are so commonly built in to other languages, this situation is worth considering.%we can retreat from trying to use the module system, but have only slightly better options.
 
@@
 %We are interested in all such pragmatic issues, so this is not an entirely satisfying solution.
-% In any case, records, tuples and labeled tuples are all library constructs in Verse.
+% In any case, records, tuples and labeled tuples are all library constructs in VerseML.
 
@@
 Refinement annotations can then be processed by external refinement checkers, like SML CIDRE \cite{davies1997refinement}.
-In Verse, we might supply refinement specifications using structured comments:
+In VerseML, we might supply refinement specifications using structured comments:
 % \begin{lstlisting}[numbers=none]
@@
 %it is inaccurate to state that they are merely syntactic sugar over such a language.
-%Object systems are perhaps even more diverse across languages\todo{citation}.
+%Object systems are perhaps even more diVerseML across languages\todo{citation}.
 % %Other dialects are justified by a desire to proivde more special-purpose semantic constructs. For example, 
@@
 
-% Verse introduces a \emph{metamodule system} that gives library providers more direct control over the type structure of the EL.
+% VerseML introduces a \emph{metamodule system} that gives library providers more direct control over the type structure of the EL.
 Using the metamodule system, library providers can express all of the types and operators discussed above.
 
-% Just as the Verse (i.e.
-ML) module system is organized around \emph{signatures} and \emph{modules}, the Verse metamodule system is organized around \emph{metasignatures} and \emph{metamodules}.
+% Just as the VerseML (i.e.
+ML) module system is organized around \emph{signatures} and \emph{modules}, the VerseML metamodule system is organized around \emph{metasignatures} and \emph{metamodules}.
 These are most easily introduced by example.
@@
 % \paragraph{Type Translations} Recall from Sec.
-\ref{sec:verse} that each external type must translate to an internal type.
+\ref{sec:VerseML} that each external type must translate to an internal type.
 Each metamodule controls the translation of types constructed by its tycons by defining a static function called the \emph{type translator}.
@@
 Metamodules like this can express many other constructs, including variants of records that support additional operations, labeled sums, object systems, constrained strings, typechecked operations on format strings like \texttt{printf} and typechecked foreign function interfaces.
-By using %This simplifies the Verse core language relative to the core languages of ML,  Scala and other comparable languages.
+By using %This simplifies the VerseML core language relative to the core languages of ML,  Scala and other comparable languages.
 
@@
 These macros also interact well with features found in typed functional languages like ML, including datatypes, modules and pattern matching.
-Consequently, Verse does not need to build in syntactic constructs that comparable languages need to (or would need to) build in, like regular expression syntax, list syntax and HTML syntax.
-% These and other examples that we will cover in the dissertation will serve to validate our claim that Verse is highly expressive (and thus should lead to a decreased need for dialect formation).
+Consequently, VerseML does not need to build in syntactic constructs that comparable languages need to (or would need to) build in, like regular expression syntax, list syntax and HTML syntax.
+% These and other examples that we will cover in the dissertation will serve to validate our claim that VerseML is highly expressive (and thus should lead to a decreased need for dialect formation).
 
 We have also outlined a formal semantics for these novel primitives.
-A more detailed specification and metatheoretic results that validate our claims that Verse has a hygienic type discipline and strong modular reasoning principles represent the main avenues for remaining work.
+A more detailed specification and metatheoretic results that validate our claims that VerseML has a hygienic type discipline and strong modular reasoning principles represent the main avenues for remaining work.
 We have already made substantial progress on closely related variants of the mechanisms proposed here, so we anticipate that the work can be completed over the remainder of the semester, per the timeline above.
@@
 % These constructs, very generally summarized, give library providers \emph{metaprogrammatic control} over aspects of the \emph{external} (i.e.
-user-facing) textual syntax and type structure of Verse, respectively.
+user-facing) textual syntax and type structure of VerseML, respectively.
 This \emph{external language} (EL) is given meaning by type-directed translation to a minimal \emph{typed internal language} (IL), specified in a completely standard manner.
 In some ways, this style of language specification is comparable to a specification for the first stage of a type-directed compiler shifted ``one level up'' into the language and modularly reorganized.
-By \emph{metaprogrammatic control}, we mean that Verse exposes a \emph{static language} (SL) to library providers.
+By \emph{metaprogrammatic control}, we mean that VerseML exposes a \emph{static language} (SL) to library providers.
 Functions written in the SL, which we refer to as \emph{static functions}, are invoked by the external semantics to control certain aspects of parsing, elaboration, typechecking and translation.
@@
 %  %while imposing constraints in order to maintain strong metatheoretic guarantees, including the guarantee that these constructs can be reasoned about modularly and used together by clients in any combination, without the possibility of syntactic or semantic conflicts.\footnote{...other than simple naming conflicts, which must be handled by adopting appropriate naming conventions, e.g.
-based on URIs as in the Java ecosystem.} By \emph{metaprogrammatic}, we mean that Verse exposes a \emph{static language} (SL) to library providers.
-Functions written in the SL, which we refer to as \emph{static functions}, are invoked by the language to control certain aspects of parsing, elaboration, typechecking and translation to Verse's minimal typed internal language (IL), which is specified in the standard manner.% In a sense, Verse pulls machinery that is usually part of a compiler into the language specification so that it can give library providers greater control over it.
+based on URIs as in the Java ecosystem.} By \emph{metaprogrammatic}, we mean that VerseML exposes a \emph{static language} (SL) to library providers.
+Functions written in the SL, which we refer to as \emph{static functions}, are invoked by the language to control certain aspects of parsing, elaboration, typechecking and translation to VerseML's minimal typed internal language (IL), which is specified in the standard manner.% In a sense, VerseML pulls machinery that is usually part of a compiler into the language specification so that it can give library providers greater control over it.
 
@@
 
-% %{\Verse} aims to explore the polar opposite design philosophy: a core language designed to remain small and as unopinionated as possible.
+% %{\VerseML} aims to explore the polar opposite design philosophy: a core language designed to remain small and as unopinionated as possible.
 Rather than building in extensive syntax and type structure \emph{a priori}, it is organized around metaprogramming mechanisms that  %In other words, power is decentralized (though not completely, as we will discuss), while responsibility over metatheoretic guarantees remains ultimately in the hands of the language designer.
@@
 % \section{Contributions}
-% \subsection{Verse}
-% To provide a coherent vehicle for our contributions, we introduce a new programming language called Verse.
+% \subsection{VerseML}
+% To provide a coherent vehicle for our contributions, we introduce a new programming language called VerseML.
 Let us briefly review its organization.
-Syntactically, Verse has a layout-sensitive textual concrete syntax (i.e.
+Syntactically, VerseML has a layout-sensitive textual concrete syntax (i.e.
 newlines and indentation are meaningful).
 This choice is not fundamental to our proposed mechanisms, but it will be useful for cleanly expressing a class of examples that we will discuss later.
-Semantically, Verse is organized into a \emph{module language} and a \emph{core language}.
+Semantically, VerseML is organized into a \emph{module language} and a \emph{core language}.
 The module language is based directly on the Standard ML module language, with which we assume familiarity for the purposes of this proposal \cite{harper1997programming}.
@@
 
-% The main novelty is that the Verse EL builds in almost no type structure or syntax \emph{a priori}.
+% The main novelty is that the VerseML EL builds in almost no type structure or syntax \emph{a priori}.
 Instead, \textbf{metamodules} and \textbf{typed syntax macros} (TSMs), described below, are expressive enough to render building in substantial type structure and concrete syntax unnecessary.
@@
 % Typed syntax macros (TSMs) allow library providers  to metaprogrammatically control the parsing and elaboration of delimited segments of concrete syntax.
-As such, Verse does not need to build in  many forms of derived syntax, e.g.
+As such, VerseML does not need to build in  many forms of derived syntax, e.g.
 list syntax as built in to many languages, but also HTML syntax, regular expression pattern  syntax and quasiquotations.
@@
 This annotation can be parameterized by types as well as modules.
-Because the Verse module system is borrowed directly from ML,  this means that TSMs can be used to define derived syntax valid for all instances of an abstract type.
-
-% To control the syntactic cost of invoking a TSM (controlling syntactic cost being, of course, the very purpose of TSMs), Verse can often locally infer these parameters.
-Moreover, when a type is declared, Verse allows the library provider to designate one TSM as its \emph{type-specific language} (TSL).
+Because the VerseML module system is borrowed directly from ML,  this means that TSMs can be used to define derived syntax valid for all instances of an abstract type.
+
+% To control the syntactic cost of invoking a TSM (controlling syntactic cost being, of course, the very purpose of TSMs), VerseML can often locally infer these parameters.
+Moreover, when a type is declared, VerseML allows the library provider to designate one TSM as its \emph{type-specific language} (TSL).
 Clients can then omit mention of this TSM entirely in positions where an expression of that type is expected.
-These mechanisms are one reason why we specify Verse by a bidirectionally typed elaboration semantics.
+These mechanisms are one reason why we specify VerseML by a bidirectionally typed elaboration semantics.
 
@@
 to the EL) of delimited segments of concrete syntax at a specified type or, perhaps most interestingly, a module-parameterized family of types.
-Consequently, Verse does not need to build in derived syntax for  constructs like lists, HTML documents, regular expressions and quasiquotations.
+Consequently, VerseML does not need to build in derived syntax for  constructs like lists, HTML documents, regular expressions and quasiquotations.
 Syntactic ambiguities are not possible by construction and, by validating the code generated by the SL, we maintain a hygienic type discipline.
@@
 % \item \textbf{Metamodules} allow library providers to express new type constructors (tycons) together with associated operators, controlling their typechecking and translation to a minimal internal language (IL) metaprogrammatically.
-As such, Verse does not need to build in constructs like tuples/records (and variants thereof), labeled sums, objects, typechecked \texttt{printf} and typechecked foreign function interfaces.
+As such, VerseML does not need to build in constructs like tuples/records (and variants thereof), labeled sums, objects, typechecked \texttt{printf} and typechecked foreign function interfaces.
 %To support tycon modules, the core language is specified like the first stage of a compiler, i.e.
@@
 This annotation can be parameterized by types as well as modules, providing a clean method for associating derived syntax with all instances of an abstract type.
-To control the syntactic cost of invoking a TSM (controlling syntactic cost being, of course, their very purpose), Verse can often locally infer these parameters.
-Moreover, when a type is declared, Verse allows the library provider to designate one TSM as its \emph{type-specific language} (TSL).
+To control the syntactic cost of invoking a TSM (controlling syntactic cost being, of course, their very purpose), VerseML can often locally infer these parameters.
+Moreover, when a type is declared, VerseML allows the library provider to designate one TSM as its \emph{type-specific language} (TSL).
 Clients can then omit mention of this TSM entirely in positions where an expression of that type is expected, further decreasing syntactic cost.
@@
 
-% %Verse provides forms of type generativity \emph{Type-specific languages} (TSLs) 
+% %VerseML provides forms of type generativity \emph{Type-specific languages} (TSLs) 
 % % clean integration with the module system.
@@
 % \end{itemize}
-% For each mechanism that we introduce, we will first demonstrate its expressive power by showing how constructs that are, or would need to be, built directly in to contemporary languages can in Verse be expressed in libraries.
+% For each mechanism that we introduce, we will first demonstrate its expressive power by showing how constructs that are, or would need to be, built directly in to contemporary languages can in VerseML be expressed in libraries.
 To demonstrate that these mechanisms are theoretically sound, we will then develop a formal specification and proofs of various important metatheoretic properties.
@@
 % % Rather than building derived syntax for certain privileged data structures, e.g.
-list syntax, as is nearly ubiquitious in dialects of ML, Verse allows users to express new derived syntax on their own, associating it with their own data structures.
-To guarantee that syntactic ambiguities cannot arise, Verse controls which delimiters can be used around derived syntax, but otherwise imposes no
+list syntax, as is nearly ubiquitious in dialects of ML, VerseML allows users to express new derived syntax on their own, associating it with their own data structures.
+To guarantee that syntactic ambiguities cannot arise, VerseML controls which delimiters can be used around derived syntax, but otherwise imposes no
 
@@
 
-%  %Constructs that are built in to languages like ML, like $n$-ary tuples/records, labeled sums and others, can in Verse be realized as modes of use of this mechanism. 
+%  %Constructs that are built in to languages like ML, like $n$-ary tuples/records, labeled sums and others, can in VerseML be realized as modes of use of this mechanism. 
 % %Languages that are organized around mechanisms like these are colloquially known as \emph{extensible languages}.
@@
 
-% %The result of this work will be the design of a \emph{modularly extensible programming language} called Verse.
+% %The result of this work will be the design of a \emph{modularly extensible programming language} called VerseML.
 and end with a brief foray into editor services in Sec.
@@
 %it is inaccurate to state that they are merely syntactic sugar over such a language.
-%Object systems are perhaps even more diverse across languages\todo{citation}.
+%Object systems are perhaps even more diVerseML across languages\todo{citation}.
 % %Other dialects are justified by a desire to provide more special-purpose semantic constructs. For example, 

### 2015-11-12 thesis-proposal/problem.tex
Updated timeline for proposal

@@
 \item The fourth premise applies the expanded parse function to the encoding of the body to produce an encoding of the expansion, $e_\text{exp}$, of  type $\mathsf{Exp}$. 
-\item Values of type $\mathsf{Exp}$ map onto \emph{marked expressions}, $\dot{e}$.
+\item The fifth premise decodes $e_\text{exp}$ to produce the \emph{marked expansion}, $\dot{e}_\text{exp}$.
+Every value of type $\mathsf{Exp}$ corresponds to a \emph{marked expressions}, $\dot{e}$.
 Per Figure \ref{fig:lambda-tsm-syntax}, marked expressions can contain variables, type variables and marked types, $\dot{\tau}$, so there is also a mapping from values of types abbreviated $\mathsf{Var}$, $\mathsf{TVar}$ and $\mathsf{Type}$ onto variables, type variables, and {marked types}, respectively.
@@
 
-The fifth premise decodes $e_\text{exp}$ to produce the \emph{marked expansion}, $\dot{e}_\text{exp}$. 
 \item The final premise valides the expansion by checking the marked expansion against the type specified by the TSM, and generates the final expansion, $e$, according to the rules defining the \emph{expansion validation judgements}:
@@
 \ref{sec:examples}.
-We leave further discussion of such a facility and of parameterized TPSMs also as remaining work (see Sec.
+We leave further discussion of such a facility and of parameterized TPSMs as remaining work (see Sec.
 \ref{sec:syntax-timeline}).
@@
 
-In the dissertation, I plan to first present the formal system $\mathcal{L}^\text{TSM}$.
+In the dissertation, I plan to first describe the simplest style of TSMs (i.e.
+unparameterized expression TSMs) by example.
+This is \textbf{complete}.
+I will next present a formal system that formalizes such TSMs.
 The account in Sec.
 \ref{sec:tsm-formalism} represents the first and most substantial step of this plan.
-This is complete on paper, so I only need to typeset the omitted details, including the metatheory.
-I plan to finish this by the \textbf{end of October}.
+This is complete on paper, so I mainly need to typeset the omitted details, including the metatheory.
+I have made substantial progress as of this writing (see link below) and I plan to finish this by the \textbf{middle of November}.
 This system will serve as the basis for several extensions that capture other constructs described above, so I will ask the committee to read this section in detail once it is complete.
 
-The first extension will add support for simple TSLs over unparameterized datatypes.
-This has already been written up in \cite{TSLs}, so I plan to have this in final form in the \textbf{first week of November}.
-
-The next extension will add support for pattern matching over these datatypes, and TPSMs.
+The first extension will add support for pattern matching.
 I will base this extension on the account of pattern matching given in \emph{PFPL} \cite{pfpl}.
-I plan to have a rough sketch of how this should work by \textbf{late-November}, in consultation with the committee, and a final write-up of it by \textbf{mid-December}.
-
-The final extension will add support for type and module parameters.
+I plan to have a rough sketch of how this should work by \textbf{late-November}, in consultation with the committee, and a final write-up by Black Friday (i.e.
+the \textbf{third week of December}).
+
+The next extension will add support for value, type and module parameters.
 To do so, I will assume that the available module paths and their signatures are being tracked by a module context in the static semantics, without detailing how the module language populates this context (i.e.
 I will only specify the ``core language'').
-I plan to sketch this out and discuss it with my committee also \textbf{during November} and finish writing the details up by the \textbf{last week of December}.
-
-Over the course of the \textbf{Spring semester}, I plan to write the remaining sections (e.g.
-additional background, related and future work sections, and additional examples), flesh out the proofs of minor lemmas and respond to the committee's requests for edits.
-Based on this timeline, I plan to defend before the \textbf{end of the Spring semester}.
-
-My current draft of the dissertation document will be available on GitHub:
+I plan to sketch this out and discuss it with my committee \textbf{during November and December} and finish writing the details up by the \textbf{beginning of Spring semester}.
+
+Beginning in the \textbf{Spring semester}, I will write up the work on type-specific languages.
+A variant of this work has been published already, and the groundwork for it will have already been laid down in the earlier chapters, so I anticipate being done with this by \textbf{the second week of February}.
+I will finish the remaining sections (e.g.
+additional background, related and future work sections, and additional examples), flesh out the proofs of minor lemmas and respond to the committee's requests for edits for the remainder of the semester.
+Based on this timeline, I plan to defend before the \textbf{end of the Spring 2016 semester}.
+
+My current draft of the dissertation document will be available and updated on an ongoing basis on GitHub:
 

### 2015-11-14 thesis-proposal/problem.tex
Updated timeline

@@
 
-\item The final premise valides the expansion by checking the marked expansion against the type specified by the TSM, and generates the final expansion, $e$, according to the rules defining the \emph{expansion validation judgements}:
+\item The final premise validates the expansion by checking the marked expansion against the type specified by the TSM, and generates the final expansion, $e$, according to the rules defining the \emph{expansion validation judgements}:
 
@@
 \ref{sec:tsm-formalism} represents the first and most substantial step of this plan.
-This is complete on paper, so I mainly need to typeset the omitted details, including the metatheory.
-I have made substantial progress as of this writing (see link below) and I plan to finish this by the \textbf{middle of November}.
+A complete draft of this chapter, with the exception of proofs at the end, is \textbf{complete}.
+I plan to finish the proofs by \textbf{Nov.
+19}.
 This system will serve as the basis for several extensions that capture other constructs described above, so I will ask the committee to read this section in detail once it is complete.
@@
 I will base this extension on the account of pattern matching given in \emph{PFPL} \cite{pfpl}.
-I plan to have a rough sketch of how this should work by \textbf{late-November}, in consultation with the committee, and a final write-up by Black Friday (i.e.
+I plan to have a rough sketch on paper of how this should work by \textbf{Dec.
+2} and a draft of the chapter (excepting complete proofs) by Black Friday (i.e.
 the \textbf{third week of December}).
 
-The next extension will add support for value, type and module parameters.
+The next extension will add support for type and module parameters.
 To do so, I will assume that the available module paths and their signatures are being tracked by a module context in the static semantics, without detailing how the module language populates this context (i.e.
@@
 
-Beginning in the \textbf{Spring semester}, I will write up the work on type-specific languages.
-A variant of this work has been published already, and the groundwork for it will have already been laid down in the earlier chapters, so I anticipate being done with this by \textbf{the second week of February}.
+Beginning in the \textbf{Spring semester}, I will be working half-time on a new grant and half-time on finishing the dissertation.
+The first task will be to write up the work on type-specific languages.
+A variant of this work has been published already, and the groundwork for it will have already been laid down in the earlier chapters, so I anticipate being done with this by \textbf{the end of February}.
 I will finish the remaining sections (e.g.
 additional background, related and future work sections, and additional examples), flesh out the proofs of minor lemmas and respond to the committee's requests for edits for the remainder of the semester.
-Based on this timeline, I plan to defend before the \textbf{end of the Spring 2016 semester}.
+Based on this timeline, I plan to have a completed draft of the dissertation by the \textbf{end of beginning of April} and defend before the \textbf{end of the Spring 2016 semester}.
 

### 2015-11-16 thesis-proposal/problem.tex
Updated timeline

@@
 
-In the dissertation, I plan to first describe the simplest style of TSMs (i.e.
-unparameterized expression TSMs) by example.
+In the dissertation, I plan to first describe the simplest style of TSMs -- unparameterized expression TSMs -- by example.
 This is \textbf{complete}.
-I will next present a formal system that formalizes such TSMs.
+I will next present a formal system that captures such TSMs.
 The account in Sec.
 \ref{sec:tsm-formalism} represents the first and most substantial step of this plan.
-A complete draft of this chapter, with the exception of proofs at the end, is \textbf{complete}.
-I plan to finish the proofs by \textbf{Nov.
-19}.
-This system will serve as the basis for several extensions that capture other constructs described above, so I will ask the committee to read this section in detail once it is complete.
+A complete draft of this chapter is \textbf{complete}.
+This system will serve as the basis for several extensions that capture other constructs described above, so I will ask the committee to read this section in detail once the proposal is complete.
 

### 2015-11-16 thesis-proposal/problem.tex
Updated formal section of proposal to be consistent with thesis

@@
 Functional programming languages come in many sizes.
-The smallest functional languages -- often referred to as ``{lambda calculi}'' -- allow researchers and language designers to study the mathematical properties of language primitives of interest in isolation.
+The smallest such languages -- often called ``{lambda calculi}'' -- allow researchers and language designers to study the mathematical properties of language primitives of interest in isolation.
 These ``small-scale'' languages inform the design of ``full-scale''
 %\footnote{Throughout this work, words and phrases that should be read as having an intuitive or informal meaning, rather than a strict mathematical meaning, will be introduced with quotation marks.} 
- languages, which combine several such primitives and also define various derived syntactic forms (colloquially, ``syntactic sugar'') that decrease the syntactic cost of selected idioms. 
+ languages, which combine several such primitives and also define various derived syntactic forms (colloquially, ``syntactic sugar'') designed to decrease the syntactic cost of selected idioms. 
 For example, Standard ML (SML) \cite{mthm97-for-dart,harper1997programming}, OCaml \cite{ocaml-manual} and Haskell \cite{jones2003haskell}   build in 
@@
 Similarly, these languages all build in
-derived forms that decrease the syntactic cost of working with lists. For example, in these languages, the form \lstinline{[1, 2, 3, 4, 5]} desugars to: 
+derived forms that decrease the syntactic cost of working with lists. In these languages, the form \lstinline{[1, 2, 3, 4, 5]} desugars to: 
 \begin{lstlisting}[numbers=none]
@@
 %Syntactic cost is often assessed qualitatively \cite{green1996usability}, though quantitative metrics can be defined.
-We will consider another example -- regular expression patterns -- in Sec. \ref{sec:examples}. 
+We will consider another example -- a dialect with syntactic support for regular expression patterns -- in Sec. \ref{sec:examples}. 
 Tools like Camlp4 \cite{ocaml-manual}, Sugar* \cite{erdweg2011sugarj,erdweg2013framework} and Racket's preprocessor \cite{Flatt:2012:CLR:2063176.2063195}, which we will discuss in Sec.
@@
 
-In some cases, coming up with a combined dialect is difficult because the dialects are specified using different formalisms.
+In some cases, coming up with such a combined dialect is difficult because the dialects are specified using different formalisms.
 In other cases, a common formalism has been used, but it does not operationalize the notion of dialect combination (e.g.
@@
 \end{itemize}
-%We will make these notions completely precise as we continue.
+We will make these notions completely precise as we continue.
 
@@
 syntax rx[Q : RX] at Q.t {
-  static fn (body : Body) : Exp => (* regex expression parser here *)
+  static fn (body : Body) : Exp option => (* regex literal parser here *)
 }
@@
 the parameter \lstinline{P} and the delimited form \lstinline{/SURLA|T|G|CEURL/} in the example above, the expansion is computed by calling the \emph{parse function} defined within braces.
-The parse function is a static function of type \lstinline{Body -> Exp}.
+The parse function is a static function of type \lstinline{Body -> Exp option}.
 Both of these types are defined in the VerseML \emph{prelude}, which is a set of definitions available ambiently.
@@
 \begin{figure}$\begin{array}{llllll}
-\textbf{variables} & \textbf{type variables} & \textbf{TSM variables} & \textbf{bodies}\\
-x & t & m & b\\~\end{array}$\\
+\textbf{variables} & \textbf{type variables} & \textbf{TSM identifiers} & \textbf{literal bodies}\\
+x & t & a & b\\~\end{array}$\\
 $\begin{array}{l}
@@
 ~\\
+\textbf{expanded expressions}\\
+e ::= x ~\vert~ \lambda x{:}{\tau}.e ~\vert~ e(e) ~\vert~ \Lambda t.e ~\vert~ e\texttt{[}\tau\texttt{]} ~\vert~ \mathsf{fold}[t.\tau](e) ~\vert~ \mathsf{unfold}(e) ~\vert~ ()\\
+~\\
+\textbf{unexpanded expressions}\\
+\hat{e} ::= {x} ~\vert~ \lambda x{:}\tau.\hat{e} ~\vert~ \hat{e}(\hat{e}) ~\vert~ \Lambda t.\hat{e} ~\vert~ \hat{e}\texttt{[}\tau\texttt{]} ~\vert~ \mathsf{fold}[t.\tau](\hat{e}) ~\vert~ \mathsf{unfold}(\hat{e}) ~\vert~ () ~\vert~ \mathsf{syntax}~a~\mathsf{at}~\tau~\{\hat{e}\} ~\mathsf{in}~\hat{e} ~\vert~ a\,\texttt{/}b\texttt{/}
+\\~\\
 \textbf{marked types}\\
@@
 ~\\
-\textbf{TSM expressions}\\
-\eta ::= m ~\vert~ \mathsf{syntax} \,@\, \tau\,\{\hat{e}\}
-\\~\\
-\textbf{unexpanded expressions}\\
-\hat{e} ::= {x} ~\vert~ \lambda x{:}\tau.\hat{e} ~\vert~ \hat{e}(\hat{e}) ~\vert~ \Lambda t.\hat{e} ~\vert~ \hat{e}\texttt{[}\tau\texttt{]} ~\vert~ \mathsf{fold}[t.\tau](\hat{e}) ~\vert~ \mathsf{unfold}(\hat{e}) ~\vert~ () ~\vert~ \mathsf{let~syntax}~m = \eta ~\mathsf{in}~\hat{e} ~\vert~ \eta\,\texttt{/}b\texttt{/}
-\\~\\
+%\textbf{TSM expressions}\\
+%\eta ::= m ~\vert~ \mathsf{syntax} \,@\, \tau\,\{\hat{e}\}
+%\\~\\
 \textbf{marked expressions}\\
@@
 \\~\\
-\textbf{expanded expressions}\\
-e ::= x ~\vert~ \lambda x{:}{\tau}.e ~\vert~ e(e) ~\vert~ \Lambda t.e ~\vert~ e\texttt{[}\tau\texttt{]} ~\vert~ \mathsf{fold}[t.\tau](e) ~\vert~ \mathsf{unfold}(e) ~\vert~ ()\end{array}\\\vspace{8px}$
-~\\
-$\begin{array}{ll}
-\textbf{type formation contexts} & \textbf{typing contexts}\\
-\Delta ::= \emptyset ~\vert~ \Delta, t & \Gamma ::= \emptyset ~\vert~ \Gamma, x : \tau
+\end{array}\\\vspace{8px}$
+$\begin{array}{lll}
+\textbf{type formation contexts} & \textbf{typing contexts} & \textbf{TSM environments}\\
+\Delta ::= \emptyset ~\vert~ \Delta, t & \Gamma ::= \emptyset ~\vert~ \Gamma, x : \tau & \Sigma ::= \emptyset ~\vert~ \Sigma, a \hookrightarrow \mathsf{syntax}(\tau; e)
 %~\\
@@
 \textbf{Judgement Form} & \textbf{Description}\\
-\Delta~\Gamma \vdash \hat{e} \leadsto e : \tau & \text{$\hat{e}$ expands to $e$ at type $\tau$ under $\Delta$ and $\Gamma$}
+\Delta~\Gamma \vdash_\Sigma  \hat{e} \leadsto e : \tau & \text{$\hat{e}$ expands to $e$ at type $\tau$ under $\Delta$, $\Gamma$ and macro environment $\Sigma$}
 \end{array}$
@@
 	\Delta \vdash \tau~\mathtt{type}\\
-	\Delta~\Gamma, x:\tau \vdash \hat{e} \leadsto e : \tau'
+	\Delta~\Gamma, x:\tau \vdash_\Sigma \hat{e} \leadsto e : \tau'
 }{
-	\Delta~\Gamma \vdash \lambda x{:}\tau.\hat{e} \leadsto \lambda x{:}\tau.e : \tau \rightharpoonup \tau'
+	\Delta~\Gamma \vdash_\Sigma \lambda x{:}\tau.\hat{e} \leadsto \lambda x{:}\tau.e : \tau \rightharpoonup \tau'
 }
@@
 \inferrule[T-U-ap]{
-	\Delta~\Gamma \vdash \hat{e}_1 \leadsto e_1 : \tau \rightharpoonup \tau' \\
-	\Delta~\Gamma \vdash \hat{e}_2 \leadsto e_2 : \tau
+	\Delta~\Gamma \vdash_\Sigma \hat{e}_1 \leadsto e_1 : \tau \rightharpoonup \tau' \\
+	\Delta~\Gamma \vdash_\Sigma \hat{e}_2 \leadsto e_2 : \tau
 }{
-	\Delta~\Gamma \vdash \hat{e}_1(\hat{e}_2) \leadsto e_1(e_2) : \tau'
+	\Delta~\Gamma \vdash_\Sigma \hat{e}_1(\hat{e}_2) \leadsto e_1(e_2) : \tau'
 }
@@
 There are two forms in the syntax of $\hat{e}$ that have no corresponding form in the syntax of $e$.
-The first allows the programmer to bind a \emph{TSM variable}, $m$, to a \emph{TSM expression}, $\eta$.
-TSM expressions are either TSM variables or \emph{TSM definitions}.
-Substitution for TSM variables is performed statically, so we only need a rule for the case where the TSM variable is being bound to a TSM definition:
+The first allows the programmer to introduce a new TSM, identified as $a$, at type $\tau$, with \emph{unexpanded parse function} $\hat{e}_\text{parse}$, into the macro environment of another unexpanded expression, $\hat{e}$:
 \begin{mathpar}
 \inferrule[T-U-TSM-let]{
-	\Delta \vdash \mathsf{syntax}~@~\tau~\{\hat{e}_\text{parse}\}~\mathtt{tsm}\\
-	\Delta~\Gamma \vdash [\mathsf{syntax}~@~\tau~\{\hat{e}_\text{parse}\}/m]\hat{e} \leadsto e : \tau'
+	\Delta \vdash \tau~\mathtt{type}\\
+	\emptyset~\emptyset \vdash_{\emptyset} \hat{e}_\text{parse} \leadsto e_\text{parse} : \mathsf{Body} \rightharpoonup \mathsf{ExpOption}\\
+	a \notin \text{dom}(\Sigma)\\
+	\Delta~\Gamma \vdash_{\Sigma, a \hookrightarrow \mathsf{syntax}(\tau; e_\text{parse})} \hat{e} \leadsto e : \tau'
 }{
-	\Delta~\Gamma \vdash \mathsf{let}~\mathsf{syntax}~m=\mathsf{syntax}~@~\tau~\{\hat{e}_\text{parse}\}~\mathsf{in}~\hat{e} \leadsto e : \tau'
+	\Delta~\Gamma \vdash_\Sigma \mathsf{syntax}~a~\mathsf{at}~\tau~\{\hat{e}_\text{parse}\}~\mathsf{in}~\hat{e} \leadsto e : \tau'
 }
 \end{mathpar}
-The first premise checks that the TSM definition is valid. It is defined by the following rule:
-\begin{mathpar}
-\inferrule[TSM-OK]{
-	\Delta \vdash \tau~\mathtt{type}\\
-	\emptyset~\emptyset \vdash \hat{e}_\text{parse} \leadsto e_\text{parse} : \mathsf{Body} \rightharpoonup \mathsf{ExpOption}
-}{
-	\Delta \vdash \mathsf{syntax}~@~\tau~\{\hat{e}_\text{parse}\}~\mathtt{tsm}
-}
-\end{mathpar}
-The first premise of (TSM-OK) checks that the type is valid.
-The second premise typechecks and expands the parse function, which must be closed.
-We discuss the abbreviated types $\mathsf{Body}$ and $\mathsf{Exp}$ below.
-
-The second form in the syntax of $\hat{e}$ that has no corresponding form in the syntax of $e$ is the form for TSM application to a delimited body, $\eta\,\texttt{/}b\texttt{/}$.
-Again because substitution for TSM variables is performed statically, we only need a rule for the case where $\eta$ is a TSM definition:
+The first premise of this rule checks that the type annotation is valid.
+The second premise types that unexpanded parse function and generates an expanded parse function, $e_\text{parse}$.
+We describe the types $\mathsf{Body}$ and $\mathsf{ExpOption}$ below.
+The third premise checks that another macro identified as $a$ is not already in scope.
+The final premise adds the definition of $a$ to the macro environment, $\Sigma$, and proceeds to type and expand $\hat{e}$.
+
+The second form in the syntax of $\hat{e}$ that has no corresponding form in the syntax of $e$ is the form for applying a TSM identified as $a$ to a delimited body: $a\,\texttt{/}b\texttt{/}$.
 \begin{mathpar}
 \inferrule[T-U-TSM-ap]{
-	\Delta \vdash \tau~\mathtt{type}\\
-	\emptyset~\emptyset \vdash \hat{e}_\text{parse} \leadsto e_\text{parse} : \mathsf{Body} \rightharpoonup \mathsf{ExpOption}\\\\
 	b \downarrow e_\text{body} : \mathsf{Body}\\
@@
 	e_\text{exp} : \mathsf{Exp} \uparrow \dot{e}_\text{exp}\\\\
-	\Delta~\Gamma; \emptyset~\emptyset \vdash \dot{e}_\text{exp} \leadsto e : \tau
+	\emptyset~\emptyset \vdash^{\Delta; \Gamma; \Sigma, a \hookrightarrow \mathsf{syntax}(\tau; e_\text{parse})} \dot{e}_\text{exp} \leadsto e : \tau
 }{
-	\Delta~\Gamma \vdash \mathsf{syntax}~@~\tau~\{\hat{e}_\text{parse}\}~\texttt{/}b\texttt{/} \leadsto e : \tau
+	\Delta~\Gamma \vdash_{\Sigma, a \hookrightarrow \mathsf{syntax}(\tau; e_\text{parse})} a\,\texttt{/}b\texttt{/} \leadsto e : \tau
 }
@@
 \begin{enumerate}
-\item The first premise checks that the type specified by the TSM is valid.
-\item The second premise typechecks and expands the parse function.
-\item The third premise encodes the body, $b$, as a term, $e_\text{body}$ of type $\mathsf{Body}$ (we will give more details on how bodies are encoded in the dissertation, but omit the definition of $\mathsf{Body}$ here for concision).
-\item The fourth premise applies the expanded parse function to the encoding of the body to produce an encoding of the expansion, $e_\text{exp}$, of  type $\mathsf{Exp}$. 
-\item The fifth premise decodes $e_\text{exp}$ to produce the \emph{marked expansion}, $\dot{e}_\text{exp}$.
+\item The first premise encodes the body, $b$, as a term, $e_\text{body}$ of type $\mathsf{Body}$ (we will give more details on how bodies are encoded in the dissertation, but omit the definition of $\mathsf{Body}$ here for concision).
+\item The second premise applies the expanded parse function to the encoding of the body to produce an encoding of the expansion, $e_\text{exp}$, of  type $\mathsf{Exp}$. 
+\item The third premise decodes $e_\text{exp}$ to produce the \emph{marked expansion}, $\dot{e}_\text{exp}$.
 Every value of type $\mathsf{Exp}$ corresponds to a \emph{marked expressions}, $\dot{e}$.
@@
 \textbf{Judgement Form} & \textbf{Description}\\
-\Delta_\text{out}; \Delta \vdash \dot{\tau} \leadsto \tau~\mathtt{type} & \text{Marked type $\dot{\tau}$ expands to $\tau$ under outer context $\Delta_\text{out}$}\\
-& \text{and current context $\Delta$.}\\
-\Delta_\text{out}~\Gamma_\text{out}; \Delta~\Gamma \vdash \dot{e} \leadsto e : \tau & \text{Marked expression $\dot{e}$ expands to $e$ at type $\tau$ under outer}\\
-& \text{contexts $\Delta_\text{out}$ and $\Gamma_\text{out}$ and current contexts $\Delta$ and $\Gamma$.}
+\Delta \vdash^{\Delta_\text{S}} \dot{\tau} \leadsto \tau~\mathtt{type} & \text{Marked type $\dot{\tau}$ expands to $\tau$ under application site}\\
+& \text{context $\Delta_\text{S}$ and current context $\Delta$.}\\
+\Delta~\Gamma \vdash^{\Delta_\text{S}; \Gamma_\text{S}; \Sigma_\text{S}} \dot{e} \leadsto e : \tau & \text{Marked expression $\dot{e}$ expands to $e$ at type $\tau$ under}\\
+& \text{application site contexts $\Delta_\text{S}$, $\Gamma_\text{S}$ and $\Sigma_\text{S}$ and current}\\
+& \text{contexts $\Delta$ and $\Gamma$.}
 \end{array}$
@@
 \begin{mathpar}
-\inferrule[T-M-var]{ }{\Delta_\text{out}~\Gamma_\text{out}; \Delta~\Gamma, x : \tau \vdash x \leadsto x : \tau}
+\inferrule[T-M-var]{ }{\Delta~\Gamma, x : \tau \vdash^{\Delta_\text{S}; \Gamma_\text{S}; \Sigma_\text{S}} x \leadsto x : \tau}
 
 \inferrule[T-M-abs]{
-	\Delta_\text{out}; \Delta \vdash \dot{\tau} \leadsto \tau ~\mathtt{type}\\
-	\Delta_\text{out}~\Gamma_\text{out}; \Delta~\Gamma, x : \tau \vdash \dot{e} \leadsto e : \tau'
+	\Delta \vdash^{\Delta_\text{S}} \dot{\tau} \leadsto \tau ~\mathtt{type}\\
+	\Delta~\Gamma, x : \tau \vdash^{\Delta_\text{S}; \Gamma_\text{S}; \Sigma_\text{S}} \dot{e} \leadsto e : \tau'
 }{
-	\Delta_\text{out}~\Gamma_\text{out}; \Delta~\Gamma \vdash \lambda x{:}\dot{\tau}.\dot{e} \leadsto \lambda x{:}\tau.e : \tau \rightharpoonup \tau'
+	\Delta~\Gamma \vdash^{\Delta_\text{S}; \Gamma_\text{S}; \Sigma_\text{S}} \lambda x{:}\dot{\tau}.\dot{e} \leadsto \lambda x{:}\tau.e : \tau \rightharpoonup \tau'
 }
@@
 \inferrule[T-M-app]{
-	\Delta_\text{out}~\Gamma_\text{out}; \Delta~\Gamma \vdash \dot{e}_1 \leadsto e_1 : \tau \rightharpoonup \tau'\\
-	\Delta_\text{out}~\Gamma_\text{out}; \Delta~\Gamma \vdash \dot{e}_2 \leadsto e_2 : \tau
+	\Delta~\Gamma \vdash^{\Delta_\text{S}; \Gamma_\text{S}; \Sigma_\text{S}} \dot{e}_1 \leadsto e_1 : \tau \rightharpoonup \tau'\\
+	\Delta~\Gamma \vdash^{\Delta_\text{S}; \Gamma_\text{S}; \Sigma_\text{S}} \dot{e}_2 \leadsto e_2 : \tau
 }{
-	\Delta_\text{out}~\Gamma_\text{out}; \Delta~\Gamma \vdash \dot{e}_1(\dot{e}_2) \leadsto e_1(e_2) : \tau'
+	\Delta~\Gamma \vdash^{\Delta_\text{S}; \Gamma_\text{S}; \Sigma_\text{S}} \dot{e}_1(\dot{e}_2) \leadsto e_1(e_2) : \tau'
 }
@@
 
-The purpose of the outer contexts is to ``remember'' the context that the macro application appeared in so that spliced subexpressions extracted from the body, which are marked with the form $\mathsf{spliced}(\hat{e})$, can be checked appropriately:
+The purpose of the application site contexts is to ``remember'' the context that the macro application appeared in so that spliced subexpressions extracted from the body, which are marked with the form $\mathsf{spliced}(\hat{e})$, can be checked appropriately:
 \begin{mathpar}
+\inferrule[T-F-spliced]{
+	\Delta_\text{S} \vdash \tau~\mathtt{type}
+}{
+	\Delta \vdash^{\Delta_\text{S}} \mathsf{spliced}(\tau) \leadsto \tau~\mathtt{type}
+}
+
 \inferrule[T-M-spliced]{
-	\Delta_\text{out}~\Gamma_\text{out} \vdash \hat{e} \leadsto e : \tau
+	\Delta_\text{S}~\Gamma_\text{S} \vdash_{\Sigma_\text{S}} \hat{e} \leadsto e : \tau
 }{
-	\Delta_\text{out}~\Gamma_\text{out}; \Delta~\Gamma \vdash \mathsf{spliced}(\hat{e}) \leadsto e : \tau
+	\Delta~\Gamma \vdash^{\Delta_\text{S}; \Gamma_\text{S}; \Sigma_\text{S}} \mathsf{spliced}(\hat{e}) \leadsto e : \tau
 }

### 2016-05-19 typy-2016/omar-026.tex
added initial abstract + intro + outline + dump for typy paper

--- 2014-10-14 atlang-modularity15/atlang-modularity15.tex
@@
 %\documentclass[12pt]{article}
-\documentclass[9pt,preprint]{sigplanconf}
+\documentclass[9pt]{sigplanconf}
 \newcommand{\sigmat}[1]{\sigma_{\text{#1}}}
@@
 
-\conferenceinfo{-}{-} 
-\copyrightyear{-} 
-\copyrightdata{[to be supplied]}
-
-\title{Extensible Gradual Typing Inside a Python}
+\conferenceinfo{MODULARITY'15,}{March 16--19, 2015, Fort Collins, CO, USA}
+\CopyrightYear{2015}
+\crdata{978-1-4503-3249-1/15/03}
+
+\title{Growing a Static Type System Inside a Python}
 %\subtitle{Unambiguously Growing a Static Type System Inside a Python}
@@
 %abstractions, all within libraries.
-Programmers prefer using abstractions available as libraries for the language they already use over those requiring a switch to a new language.
-But library providers cannot typically extend the static type systems of these languages directly, so powerful statically typed abstractions  remain obscure.
-This suggests the need for a mechanism that allows abstraction providers to extend the type system of a widely adopted language from within, in a way that guarantees that any combination of extensions can be used together without worrying about conflicts.
-
-Our aim in this paper is to 1) introduce such a mechanism; and 2) show how it can be implemented as a library for an existing widely used language, here Python.
-We describe two artifacts: @\texttt{lang}, an  extensible statically typed language implemented as a library within Python, and @$\lambda$, a minimal core calculus.
-Both @\texttt{lang} and @$\lambda$ have a fixed syntax.
-They determine the semantics of each term by delegating control, according to a simple but powerful bidirectionally typed elaboration semantics, to methods associated with a relevant \emph{active type constructor}, defined in a library.
-This process is stable even when additional extensions are imported, so we avoid syntactic and semantic ambiguity by construction.
-We validate our approach with several examples of interesting type system fragments that are built into other languages but can be implemented as libraries using our approach, including functional constructs, a type system for secure string sanitation and a typed foreign function interface to OpenCL.% By flexibly  repurposing a fixed syntax in this type-directed manner, we avoid ambiguities by construction.
+Programmers prefer using constructs available as libraries for the programming language they already use over those requiring a switch to a new language.
+But library providers have not been able to modularly extend the type systems of these languages, so some of the most powerful constructs have remained obscure.
+%Our aim in this paper is to introduce a mechanism that allows abstraction providers to extend the type system of a language from within, in a way that guarantees that any combination of extensions can be used together without worrying about conflicts.
+Our aims in this paper are to 1) introduce a mechanism that allows library providers to define new type system fragments modularly; 2) show how it can itself be implemented as a library, \texttt{typy}, for a widely used language, Python; 3) validate our approach with examples of non-trivial type system fragments that are built into other languages but that can be implemented as libraries using this mechanism, including functional constructs, an object system, a type system for secure string sanitation and a typed foreign function interface to OpenCL; and 4) formally specify the mechanism with a minimal typed lambda calculus, @$\lambda$.
+The key idea is to associate semantic logic with type constructors, rather than syntactic forms.
+%Both \texttt{typy} and @$\lambda$ have a fixed syntax. 
+%The semantics of each term is determined by delegating control, during the process of typechecking and translating the term to the underlying \emph{internal language}, to methods associated with a relevant \emph{active type constructor}, defined in a library.
+This process is stable even as additional type constructors are imported, so ambiguities cannot arise.
+% By flexibly  repurposing a fixed syntax in this type-directed manner, we avoid ambiguities by construction.
 %In @\texttt{lang}, types and functions are defined using Python's syntax, distributed using Python's package system and can call into Python libraries, so @\texttt{lang} is uniquely backwards compatible.
@@
 
-%\category{D.3.2}{Programming Languages}{Language Classifications}[Extensible Languages]
+\category{D.3.2}{Programming Lan\-guages}{Language Classifications}[Extensible Languages]
 %\category{D.3.4}{Programming Languages}{Processors}[Compilers]
 %\category{F.3.1}{Logics \& Meanings of Programs}{Specifying and Verifying and Reasoning about Programs}[Specification Techniques]
-%\keywords
-%type-level computation, typed compilation
+\keywords
+type systems; extensible languages; metaprogramming
 \section{Introduction}\label{intro}
-Asking programmers to import a library is easier than asking them to adopt a new programming language.
+Asking programmers to adopt a library is easier than asking them to adopt a new programming language.
 Indeed, recent empirical studies underscore the difficulties of driving new languages into adoption, finding that extrinsic factors like compatibility with large existing code bases, library availability, familiarity and tool support are at least as important as intrinsic features of the language \cite{Meyerovich:2013:EAP:2509136.2509515,chen05}.
 
-Unfortunately, researchers and domain experts designing new abstractions sometimes find it difficult to soundly, completely and naturally implement them as libraries, particularly those that rely on specialized support from a static type system.
-In these situations, abstraction providers generally develop a new language dialect, increasingly with the aid of tools like language workbenches \cite{erdweg2013state}.
-Unfortunately, this \emph{language-oriented approach} \cite{journals/stp/Ward94} is problematic at language boundaries: working with components written in several distinct languages often requires programming against their low-level implementation details, rather than their semantics, via a foreign-function interface (FFI) to a common implementation language (e.g.
-C or the JVM).
-This is brittle, unnatural and unsafe and can cause performance issues due to necessary checks and conversions at language boundaries.
-As more languages are used together, more FFIs need to be constructed and traversed.%To avoid this, it is necessary to combine the different languages and their associated tools, but this often requires significant effort and expertise (we will discuss this further in Sec.
+Yet researchers and domain experts continue to design new languages and language dialects, justifying this practice by  arguing that it would be  impossible to isomorphically express the constructs they need as libraries for existing languages.
+This is certainly a convincing argument when the constructs in question specify a non-trivial static semantics.
+%This practice is encouraged by the increasing availability of tools like language workbenches \cite{erdweg2013state}.
+Unfortunately, the flexibility afforded taking such a \emph{language-oriented approach} \cite{journals/stp/Ward94} to new ideas comes at the cost of modularity: working with a component written in a different language typically requires programming against the low-level details of a particular implementation of that language, rather than its semantics, via a foreign-function interface (FFI) (e.g.
+to C, or JVM bytecode).
+This is brittle (the implementation may change), unnatural and unsafe.
+Mitigating the loss of static safety by adding run-time checks at language boundaries has a performance cost.
+%As more languages are used together, more FFIs need to be constructed and traversed.% This can defeat the productivity gains afforded by these dialects.%To avoid this, it is necessary to combine the different languages and their associated tools, but this often requires significant effort and expertise (we will discuss this further in Sec.
 \ref{related-work}).
 
-These issues prevent even those dialects that introduce only a few new constructs from being adopted.
-A study comparing a Java dialect, Habanero-Java (HJ), with a comparable library, \lstinlinep{java.util.concurrent}, for example, found that the language-based abstractions in HJ were easier to use and provided useful static guarantees \cite{cave2010comparing}.
+These problems are relevant even to dialects that introduce only a few new constructs.
+A study comparing a Java dialect, Habanero-Java (HJ), with a  library, \lstinlinep{java.util.concurrent}, found that the language-based abstractions in HJ were indeed easier to use and provided useful static guarantees \cite{cave2010comparing}.
 Nevertheless, it concluded that the library-based abstractions remained more practical outside the classroom because using HJ, as a distinct dialect of Java with its own type system, would be problematic in settings where some developers had not adopted it, but needed to interface with code that had.
@@
 Moreover, its tool support is more limited.
-This appears to be a widespread problem: programmers and development teams often cannot use  the typed abstractions they prefer because they are only available bundled with specialized languages they cannot  adopt \cite{Meyerovich:2012:SDR:2414721.2414724,Meyerovich:2013:EAP:2509136.2509515}.
+This appears to be a widespread problem: programmers and development teams often cannot use  the constructs they might prefer because they are only available in specialized dialects they cannot  adopt \cite{Meyerovich:2012:SDR:2414721.2414724,Meyerovich:2013:EAP:2509136.2509515}.
 %These languages become, in essence, merely vehicles for communicating ideas, rather than useful tools.
@@
 
-Extensible languages have promised to reduce the problems associated with distinct dialects by giving abstraction providers more direct and granular control over a language's syntax and semantics from within, or alongside, libraries.
+Language extension mechanisms have long promised to reduce these problems by  giving providers the ability to define syntactic and semantic constructs more granularly, so that clients can import them as needed.
 %In   an internally extensible language, programmers  can \emph{granularly} import the primitive constructs best suited to each component of their application or library.
@@
 Unfortunately, the mechanisms available today have several problems.
-First, they are themselves often available only within a dialect of an existing language and so face the classic ``chicken-and-egg'' problem: a language like SugarJ \cite{erdweg2011sugarj} or Wyvern \cite{TSLs} (which permit only syntax extensions over a fixed semantics) must overcome the same extrinsic barriers as a language like HJ.
-Second, giving abstraction providers too much control over the language can permit the definition of conflicting  extensions, a problem that is detected ``too late'': when extensions are combined by abstraction clients, who are not necessarily  capable of understanding and fixing the conflict.% Too many constraints can leave it difficult to express real-world abstractions.%It is difficult to guarantee that extensions are sound in isolation and more so to guarantee that they can be {safely} combined.
+First, they are themselves often housed in an obscure language dialect and so face the classic ``chicken-and-egg'' problem: languages like SugarJ \cite{erdweg2011sugarj} or Wyvern \cite{TSLs} (which permit syntax extensions over a fixed semantics) must overcome the same sorts of  barriers to adoption as a language like HJ.
+Second, they give providers too much control over the language, again at the cost of modularity, because providers can define conflicting constructs.
+The conflicts, if detected at all, are detected too late: when a client attempts to use the conflicting constructs in the same program.
+%Clients are not generally capable of understanding and fixing the issue productively.% Too many constraints can leave it difficult to express real-world abstractions.%It is difficult to guarantee that extensions are sound in isolation and more so to guarantee that they can be {safely} combined.
 %This requires further research before it would be appropriate to widely rely on them (which we will discuss as we go on).
@@
 
-Our focus in this paper is to enable the introduction of new static type system fragments as safely composable libraries for an existing and widely-used language, leaving its concrete and abstract syntax fixed.
-This choice is freeing: questions of syntactic ambiguity no longer need to be addressed.
-Moreover, many tools operate by exhaustive case analysis over the concrete or abstract syntax of the terms of the language they work with (e.g.
-syntax highlighters, editor modes, pretty-printers, style checkers, documentation generators).
-These tools will continue to work.
+In this paper, we take steps toward addressing both of these problems by introducing a mechanism that allows library providers to define new type system fragments modularly, and showing how it can be itself be implemented as a library, \texttt{typy}, for an existing and widely-used language, Python (using its quasiquotation and reflection facilities).
+This constrains us: we do not have the power to change Python's concrete and abstract syntax.
+But in fact, we find that this constraint  can be freeing: questions of syntactic conflict never arise, and tools that operate by exhaustive case analysis over the syntax (e.g.
+syntax highlighters, editor plugins, pretty-printers, style checkers, documentation generators) continue to work.
 %Indeed, enabling the decentralized introduction of new term constructors is the standard example used when describing the more general problem of enabling the extension of a sum type with new cases as well as new  functions over its cases  \cite{Reynolds75}, prompting Wadler to dub it the \emph{expression problem} \cite{wadler1998expression}.
 
-\paragraph{Approach} Our key insight is that few tools operate by exhaustive case analysis over type constructors.
-We introduce a mechanism that allows library providers to introduce new type system fragments by defining only new type constructors, in libraries, while leaving the concrete and abstract syntax, and the base semantics of the language, fixed.
-
-To support new operations, the semantics of terms are no longer determined directly based on the identity of their term constructors (the usual ``syntax-directed'' approach).
+Library providers introduce type system fragments by defining new type constructors (\emph{tycons}, for brevity).
+For example, a library provider might introduce record types, as are built in to functional languages like ML and Haskell, by defining a new tycon, \lstinline{record}.
+A different library provider may introduce a statically-typed variant of an object system featuring prototypic inheritance (al\'a Javascript) by defining the tycon \lstinline{proto}.
+Types are constructed by applying a tycon to a statically valued \emph{type index}, written \lstinline{tycon[tyidx]}.
+For example, we can construct a record type with two rows like this:
+\begin{lstlisting}[numbers=none]
+record[{
+  'label1' : dyn[()],    
+  'label2' : string[()]
+}]
+\end{lstlisting}
+The index is a dictionary mapping row labels to their corresponding types.
+Here, \lstinline{dyn} (which we discuss later) and \lstinline{string} are trivially indexed tycons, i.e.
+the only valid index is \lstinline{()}.
+For concision, trivial indices can simply be omitted (we showed them here only to emphasize the uniform manner in which one should think about type construction).
+A prototypic object type having an analagous field structure (and for simplicity, assuming the default prototype) is constructed like this:
+\begin{lstlisting}[numbers=none]
+proto[{ # (fields of a prototypic object are mutable)
+  'label1': dyn,   # equivalent to dyn[()]
+  'label2': string # equivalent to string[()]
+}]
+\end{lstlisting}
+
+Given that a programmer might be using both of these libraries in the same program, how should our system determine the semantics of an expression like \lstinline{e.label1}, where \lstinline{e} is some sub-expression?
+In a monolithic language, the usual approach is to define the semantics in a \emph{syntax-directed} manner: the language specifies, in one place, all the logic relevant to each form.
+Our approach, in contrast, is \emph{tycon-directed}: the language delegates  to a \emph{statically valued function} defined by some relevant tycon, called the \emph{delegate}.
+The language specifies only the \emph{delegation protocol} that assigns a delegate to each form.
+In this example, the delegate is the tycon of the type recursively synthesized by \lstinline{e}.
+We will see how tycons like \lstinline{record} and \lstinline{proto} statically compute a type and a translation (to Python code) for such a term when they are delegated control by this protocol in Sec.
+\ref{usage}.
+For now, it suffices to point out that the code involved is almost exactly what one would write if implementing a language that built in record types directly.
+
+% A statically-valued function 
+% %, identified based on the name Python's grammar gives to the form, here \lstinline{syn_Attribute},  %So if \lstinline{e} is of record type, then the \lstinline{record} tycon  is delegated control, and if \lstinline{e} is of prototypic object type then \lstinline{proto} is delegated control, in either case by calling \lstinline{syn_Attribute} with a reference to the syntax tree of the term as well as a \emph{typing context}, which we will describe later.
+% must  either synthesize a type for the term as a whole or indicate a type error.
+If it synthesizes a type, the tycon must then generate a \emph{translation} for the term as a whole, which determines its dynamic behavior.
+So, for the term \lstinline{e.label1}, if \lstinline{e} is of record type, then the \lstinline{record} tycon is delegated control.
+It will check that \lstinline{label1} is one of the row labels specified in the index of the type of \lstinline{e}.
+Records can be implemented by translation to tuples (the row labels are only needed statically), so the translation it then generates will project out the appropriate position from the translation recursively generated for \lstinline{e}.
+
+% We will detail how tycons are implemented as Python classes in Sec.
+\ref{usage}.
+The innovation is in how this protocol decentralizes control over a fixed set of forms, allowing us to define a range of semantic constructs in libraries.
+The language itself needs to specify only the \emph{delegation protocol} that determines how a tycon is delegated control over each syntactic form.
+
+% More specifically, what this means is that the statically valued function  because that the tycon defines must either synthesize a type for the term as a whole or indicate a type error.
+If a type is synthesized, the tycon must also generate a \emph{translation} for the term as a whole.
+
+% We call a type constructor equipped with a set of such statically valued functions an \emph{active type constructor}.
+
+% More specifically, the semantics are a \emph{bidirectionally-typed type-directed translation semantics}.
+
+% allow the associated operations, the semantics of terms are no longer determined directly based on the identity of their term constructors (the usual ``syntax-directed'' approach).
 Instead, we delegate to a statically defined function associated with a relevant type constructor.
-The term constructor's identity only serves to determine how the delegate is assigned.
-We call user-defined type constructors equipped with these static functions and distributed in libraries \emph{active type constructors}.
-
-As a simple example, consider the concrete term \lstinlinep{e.r} (where \lstinlinep{e} is an expression and \lstinlinep{r} is a static label), which maps to the abstract term $\FF{attr}[\text{``r''}](e)$ (i.e.
-the term constructor $\FF{attr}$, \emph{indexed} by a static representation of the label and taking one \emph{argument}, $e$).
-A fragment defining statically typed pairs might see this term as referring to projection of the right component of the pair $e$, while a fragment defining functional record types might want it project out a field labeled \verb|r| from $e$, and an object system might introduce a more complex semantics involving dynamic dispatch.
+The term constructor's identity only serves to determine how to determine this delegate.
+For example, consider the concrete term \lstinlinep{x.r} (where \lstinlinep{x} is a variable and \lstinlinep{r} is a static label), which, according to Python's grammar, maps to the abstract term \lstinlinep{Attribute("r"; Name("x"))} (i.e.
+the term constructor $\FF{Attribute}$, \emph{indexed} by a static representation of the label and one \emph{argument}, $e$).
+There are many possible meanings for this term.
+If $e$ is a pair, this might be a way to project out its right component.
+If $e$ were a functional record,  this might be the way to project out a field labeled \verb|r|.
+An object system might introduce a more complex semantics involving dynamic dispatch.
 The type of dynamically classified Python values would follow Python's own complex field lookup protocol.
-Were we to naively ``open up'' the syntax and semantics, these fragments would conflict.
-
-Instead, our semantics will delegate responsibility for assigning a type and a translation to this term to a static function associated with the type constructor of the type recursively synthesized by $e$, the \emph{target} of the operation.
+Were we to naively ``open up'' the syntax and semantics in an attempt to allow all of these features into our language, these semantics would conflict.
+
+% Instead, our semantics will delegate responsibility for assigning a type and a translation to this term to a static function associated with the type constructor of the type recursively synthesized by $e$, the \emph{target} of the operation.
 Thus, tuples, records, objects, dynamic values and variants thereof can  share this syntax.
 
-As we will detail, terms for which there is no clear ``target'' from which to determine a delegate (e.g.
-number or list literals) instead either use the type they are being analyzed against (following the approach for composable syntax extensions as supported in Wyvern \cite{TSLs}) or, if not available, a ``default'' delegate determined by the type constructor governing the function in which they appear.
-Symmetric binary terms (e.g.
-addition) use a third protocol, which we will discuss.
-There is always exactly one active type constructor delegated  responsibility over a term and this delegate is stable with respect to additional extension imports, so semantic ambiguities cannot arise by construction (unlike systems that treat the type system as a ``bag of rules'').
+This example was of a form for which there is a clear ``target'' subexpression that can be examined recursively to determine the tycon to delegate control to.
+For other forms, it is perhaps less clear.
+For example, the curly-brace-delimited form in the following expression could be given a record type, a prototypic object type or be a dynamically tagged Python dictionary, in which case it would have type \lstinline{dyn}:
+\begin{lstlisting}[numbers=none]
+f({label1: "val1", label2: "val2"})
+\end{lstlisting}
+We resolve this ambiguity by leveraging the \emph{local type inference} mechanism that \verb|typy| uses \cite{Pierce:2000:LTI:345099.345100}.
+That is, for literal forms in Python's grammar, the tycon of the type that the term is being analyzed against is delegated control.
+Here, if \lstinline{f} is a function taking an argument of the record type defined above, then \lstinline{record} is delegated control over the curly-brace-delimited form.
+This can be understood as a variant of the mechanism that Wyvern introduced for resolving ambiguous syntax extensions \cite{TSLs}, here applied in a setting where there is a fixed syntax but an extensible semantics.% When a literal form appears at the top-level of a function, where there is no expected type, the programmer can provide a type annotation on an expression directly, using the form \lstinline{e [: T]}, where \lstinline{e} is an expression and \lstinline{T} is a type (or a tycon applied to an incomplete index, which we discuss in Sec.
+\todo{inc indices}).
+
+Symmetric binary terms (e.g. addition) use a third protocol, which we will discuss.
+
+There is always a single tycon delegated  responsibility over a term and this delegate is stable with respect to additional library imports, so semantic ambiguities cannot arise by construction (unlike systems that treat the type system as a ``bag of rules'').
 %This is a generalization of a previous mechanism that we have developed, which deals only with extending the elaboration, but not the typechecking, of the bodies of delimited literals \cite{omar-ecoop14}.
@@
 
-\paragraph{Specific Aims} Our specific aims in this paper are to introduce active type constructors and demonstrate that they are expressive, practically realizable within existing languages and theoretically well-founded.
+\paragraph{Specific Aims} Our specific aims in this paper are to introduce this approach, which we call \emph{active typechecking and translation}, and demonstrate that they are expressive, practically realizable within existing languages and theoretically well-founded.
 Our target audience is typed language fragment providers (researchers or domain experts) and language designers interested in extensibility mechanisms.
@@
 % We recommend abbrvnat bibliography style.
-\newpage
+%\newpage
 \bibliographystyle{abbrv}

### 2016-06-17 atlang-modularity15/atlang-modularity15.tex
progress on GPCE paper

@@
 %\documentclass[12pt]{article}
-\documentclass[9pt,preprint]{sigplanconf}
+\documentclass{sigplanconf}
 \newcommand{\sigmat}[1]{\sigma_{\text{#1}}}
@@
 %\setlength{\parsep}{0pt}
-\setlength{\headsep}{10pt}
+%\setlength{\headsep}{10pt}
 %\setlength{\topskip}{0pt}
@@
 
-\newcommand{\lstinlinep}[1]{\lstinline[language=Python,basicstyle=\ttfamily\small,deletendkeywords={tuple,buffer,map}]{#1}}
+\newcommand{\lip}[1]{\lstinline[language=Python,basicstyle=\ttfamily\small,deletendkeywords={tuple,buffer,map}]{#1}}
 
@@
 
-\lefthyphenmin=8
+%\lefthyphenmin=8
 \sloppy
@@
 
-\conferenceinfo{-}{-} 
-\copyrightyear{-} 
-\copyrightdata{[to be supplied]}
-
-\title{Extensible Gradual Typing Inside a Python}
+\conferenceinfo{MODULARITY'15,}{March 16--19, 2015, Fort Collins, CO, USA}
+\CopyrightYear{2015}
+\crdata{978-1-4503-3249-1/15/03}
+
+\title{{Growing} a Static Type System Inside a Python}
 %\subtitle{Unambiguously Growing a Static Type System Inside a Python}
@@
 \authorinfo{Cyrus Omar \and Jonathan Aldrich}
-          {          Carnegie Mellon University}
-          {\{comar, aldrich\}@cs.cmu.edu}
+          {          Carnegie Mellon University\\
+          Pittsburgh, PA, USA}
+          {\{comar, aldrich\}@cs.cmu.edu\vspace{-10px}}
 
@@
 %
-%We introduce @\texttt{lang}, a language compatible with the popular Python ecosystem.
+%We introduce \texttt{typy}, a language compatible with the popular Python ecosystem.
 %While Python, like other similar languages, is satisfactory for simple
-%scripting, @\texttt{lang} is designed for more complex situations where static
-%typechecking is beneficial. Unlike most statically-typed languages, @\texttt{lang}'s type
+%scripting, \texttt{typy} is designed for more complex situations where static
+%typechecking is beneficial. Unlike most statically-typed languages, \texttt{typy}'s type
 %system can be extended from within, by associating compile-time functions with
@@
 %type-directed protocol, avoiding interference problems. Despite these design
-%constraints, active types in @\texttt{lang} can express the static and dynamic semantics
+%constraints, active types in \texttt{typy} can express the static and dynamic semantics
 %of a range of functional, object-oriented, parallel and domain-specific
 %abstractions, all within libraries.
-Programmers prefer using abstractions available as libraries for the language they already use over those requiring a switch to a new language.
-But library providers cannot typically extend the static type systems of these languages directly, so powerful statically typed abstractions  remain obscure.
-This suggests the need for a mechanism that allows abstraction providers to extend the type system of a widely adopted language from within, in a way that guarantees that any combination of extensions can be used together without worrying about conflicts.
-
-Our aim in this paper is to 1) introduce such a mechanism; and 2) show how it can be implemented as a library for an existing widely used language, here Python.
-We describe two artifacts: @\texttt{lang}, an  extensible statically typed language implemented as a library within Python, and @$\lambda$, a minimal core calculus.
-Both @\texttt{lang} and @$\lambda$ have a fixed syntax.
-They determine the semantics of each term by delegating control, according to a simple but powerful bidirectionally typed elaboration semantics, to methods associated with a relevant \emph{active type constructor}, defined in a library.
-This process is stable even when additional extensions are imported, so we avoid syntactic and semantic ambiguity by construction.
-We validate our approach with several examples of interesting type system fragments that are built into other languages but can be implemented as libraries using our approach, including functional constructs, a type system for secure string sanitation and a typed foreign function interface to OpenCL.% By flexibly  repurposing a fixed syntax in this type-directed manner, we avoid ambiguities by construction.
-%In @\texttt{lang}, types and functions are defined using Python's syntax, distributed using Python's package system and can call into Python libraries, so @\texttt{lang} is uniquely backwards compatible.
-We show that active type constructors as implemented in @\texttt{lang} permit practical library implementations of a variety of type system fragments that would normally need to be built into a language.
-%We also show how to construct safe, natural and extensible foreign function interfaces using a novel form of ``staged'' type propagation from Python into @\texttt{lang} and then through to OpenCL.
+Programmers prefer using constructs available as libraries for the programming language they already use over those requiring that they switch to a new language dialect.
+Unfortunately, dialects are seen as the only viable vehicle for constructs that specify non-trivial type systems.
+%Our aim in this paper is to introduce a mechanism that allows abstraction providers to extend the type system of a language from within, in a way that guarantees that any combination of extensions can be used together without worrying about conflicts.
+Our aim in this paper is to show that there is another way.
+We 1) introduce a mechanism that gives library providers the ability to introduce new type system fragments modularly; 2) show how this mechanism can itself be implemented as a library, \texttt{typy}, for an existing language, Python; 3) validate our approach with examples of non-trivial type system fragments that are built into other languages but that can be implemented as libraries using this mechanism;
+%, including functional constructs, an object system, a type system for secure string sanitation and a typed foreign function interface to OpenCL
+and 4) formally specify the mechanism with a minimal typed lambda calculus, @$\lambda$.
+The key idea is simple: rather than defining the semantics in the usual syntax-directed manner, we delegate control over the typechecking and translation of each term to some contextually-relevant type constructor.
+Users  define new type constructors in libraries.
+%We call type constructors bundled with statically valued functions of this sort \emph{active type constructors}, and this protocol \emph{active typechecking and translation}.%The key idea is to associate semantic logic with type constructors, rather than syntactic forms.
+%Both \texttt{typy} and @$\lambda$ have a fixed syntax. 
+%The semantics of each term is determined by delegating control, during the process of typechecking and translating the term to the underlying \emph{internal language}, to methods associated with a relevant \emph{active type constructor}, defined in a library.
+This process is stable even as additional type constructors are imported, so ambiguities cannot arise.
+% By flexibly  repurposing a fixed syntax in this type-directed manner, we avoid ambiguities by construction.
+%In \texttt{typy}, types and functions are defined using Python's syntax, distributed using Python's package system and can call into Python libraries, so \texttt{typy} is uniquely backwards compatible.
+We show that active type constructors as implemented in \texttt{typy} permit practical library implementations of a variety of type system fragments that would normally need to be built into a language.
+%We also show how to construct safe, natural and extensible foreign function interfaces using a novel form of ``staged'' type propagation from Python into \texttt{typy} and then through to OpenCL.
 % can be used to enable type system extensions and demonstrates how type safety can be maintained by lifting a technique from the typed compilation literature into the language.%To show that the method scales, we give a full implementation of the OpenCL programming language as a library.
 %We briefly examine the type theoretic foundations of this approach and give modular criteria that guarantee that an extension is safe and modular.
-%@\texttt{lang} can be used from the shell or interactively from within Python\todo{take this sentence out?}.
+%\texttt{typy} can be used from the shell or interactively from within Python\todo{take this sentence out?}.
 %Researchers developing languages and abstractions for high-performance computing must consider a number of design criteria, including performance, verifiability, portability and ease-of-use.
@@
 We argue that this can be largely attributed to a failure to consider three additional criteria: continuity, extensibility\- and interoperability.
-This paper introduces @\texttt{lang}, a language that aims to satisfy this more comprehensive set of design criteria.
-To do so, @\texttt{lang} introduces several novel compile-time mechanisms, makes principled design choices, and builds upon existing standards in HPC, particularly Python and OpenCL.
+This paper introduces \texttt{typy}, a language that aims to satisfy this more comprehensive set of design criteria.
+To do so, \texttt{typy} introduces several novel compile-time mechanisms, makes principled design choices, and builds upon existing standards in HPC, particularly Python and OpenCL.
 OpenCL support, rather than being built into the language, is implemented atop an extensibility mechanism that also admits abstractions drawn from other seemingly disparate paradigms.
-The core innovation underlying this and other features of @\texttt{lang} is a novel reification of types as first-class objects at compile-time, representing a refinement to the concept of active libraries that we call \emph{active types}.
+The core innovation underlying this and other features of \texttt{typy} is a novel reification of types as first-class objects at compile-time, representing a refinement to the concept of active libraries that we call \emph{active types}.
 We validate our overall design by considering a case study of a simulation framework enabling the modular specification and efficient execution of ensembles of neural simulations across a cluster of GPUs.
@@
 
-%\category{D.3.2}{Programming Languages}{Language Classifications}[Extensible Languages]
+\category{D.3.2}{Programming Lan\-guages}{Language Classifications}[Extensible Languages]
 %\category{D.3.4}{Programming Languages}{Processors}[Compilers]
 %\category{F.3.1}{Logics \& Meanings of Programs}{Specifying and Verifying and Reasoning about Programs}[Specification Techniques]
-%\keywords
-%type-level computation, typed compilation
+\keywords
+type systems; extensible languages; metaprogramming
 \section{Introduction}\label{intro}
-Asking programmers to import a library is easier than asking them to adopt a new programming language.
-Indeed, recent empirical studies underscore the difficulties of driving new languages into adoption, finding that extrinsic factors like compatibility with large existing code bases, library availability, familiarity and tool support are at least as important as intrinsic features of the language \cite{Meyerovich:2013:EAP:2509136.2509515,chen05}.
-
-Unfortunately, researchers and domain experts designing new abstractions sometimes find it difficult to soundly, completely and naturally implement them as libraries, particularly those that rely on specialized support from a static type system.
-In these situations, abstraction providers generally develop a new language dialect, increasingly with the aid of tools like language workbenches \cite{erdweg2013state}.
-Unfortunately, this \emph{language-oriented approach} \cite{journals/stp/Ward94} is problematic at language boundaries: working with components written in several distinct languages often requires programming against their low-level implementation details, rather than their semantics, via a foreign-function interface (FFI) to a common implementation language (e.g.
-C or the JVM).
-This is brittle, unnatural and unsafe and can cause performance issues due to necessary checks and conversions at language boundaries.
-As more languages are used together, more FFIs need to be constructed and traversed.%To avoid this, it is necessary to combine the different languages and their associated tools, but this often requires significant effort and expertise (we will discuss this further in Sec.
+Convincing programmers to adopt constructs distributed in libraries is easier than con\-vinc\-ing them to use a new programming language.
+Indeed, re\-cent empirical studies underscore the  difficulty of driving language adoption, finding that extrinsic factors like library compatibility, familiarity and tool support are at least as important to programmers as the intrinsic features of a language \cite{Meyerovich:2013:EAP:2509136.2509515,chen05}.
+
+Despite these steep barriers to adoption, researchers and domain experts continue to design new language dialects, arguing that they have no choice: there often seems to be no other way to faithfully express the constructs that they need, particularly those  specified by a non-trivial static type system.
+As a concession to concerns about library compatibility, these dialects will often build in a foreign function interface (FFI) to a popular language, like Java.
+%\emph{Language workbenches} \cite{erdweg2013state} have made it easy to develop sophisticated tooling around such dialects.
+%As a result, there are now a large number of dialects structured in this manner.
+
+This approach solves the important problem of interfacing with the libraries already written in  this popular language.
+Interfacing with a library written in a sibling  dialect, however, is difficult.
+To access such a library via the FFI requires programming against that dialect's implementation, i.e.
+the code generated by its compiler, rather than its semantics directly.
+This is anti-modular (if the dialect implementation changes, client code will break), unnatural (syntactic conveniences provided by the dialect are not available) and, for  dialects featuring a type system that statically enforces non-trivial invariants, unsafe.
+This safety threat can at best be mitigated by including run-time checks in anticipation of this mode of use, though this requires additional effort and comes at the cost of performance.% This likely explains much of the resistance that developers have toward ideas housed in ``toy dialects''.
+%As more languages are used together, more FFIs need to be constructed and traversed.% This can defeat the productivity gains afforded by these dialects.%To avoid this, it is necessary to combine the different languages and their associated tools, but this often requires significant effort and expertise (we will discuss this further in Sec.
 \ref{related-work}).
 
-These issues prevent even those dialects that introduce only a few new constructs from being adopted.
-A study comparing a Java dialect, Habanero-Java (HJ), with a comparable library, \lstinlinep{java.util.concurrent}, for example, found that the language-based abstractions in HJ were easier to use and provided useful static guarantees \cite{cave2010comparing}.
-Nevertheless, it concluded that the library-based abstractions remained more practical outside the classroom because using HJ, as a distinct dialect of Java with its own type system, would be problematic in settings where some developers had not adopted it, but needed to interface with code that had.
-It would also be difficult to use it in combination with other abstractions also implemented as dialects of Java.
-Moreover, its tool support is more limited.
-This appears to be a widespread problem: programmers and development teams often cannot use  the typed abstractions they prefer because they are only available bundled with specialized languages they cannot  adopt \cite{Meyerovich:2012:SDR:2414721.2414724,Meyerovich:2013:EAP:2509136.2509515}.
+These problems are severe enough to render even dialects that conservatively introduce only a few new constructs impractical.
+For example, an empirical study comparing a Java dialect, Habanero-Java (HJ), with a library, \lip{java.util.concurrent}, found that the constructs in HJ were indeed easier to use and provided useful static guarantees \cite{cave2010comparing}.
+Nevertheless, it concluded that the library-based constructs remained more practical outside the classroom because using HJ, as a distinct dialect of Java with its own type system, would be problematic in settings where some developers had not adopted it.
+It would also be difficult to use it in combination with other constructs also implemented in dialects of Java.
+%Moreover, its tool support is more limited.
+
 %These languages become, in essence, merely vehicles for communicating ideas, rather than useful tools.
@@
 
-Extensible languages have promised to reduce the problems associated with distinct dialects by giving abstraction providers more direct and granular control over a language's syntax and semantics from within, or alongside, libraries.
+Language extension mechanisms have promised to mitigate the need for distinct dialects by giving providers the ability to define new syntactic and semantic constructs more granularly, allowing library clients to import them wherever needed.
 %In   an internally extensible language, programmers  can \emph{granularly} import the primitive constructs best suited to each component of their application or library.
@@
 %Indeed, abstractions could compete on their individual merit, rather than  by the serious adoption barriers mentioned above.
-Unfortunately, the mechanisms available today have several problems.
-First, they are themselves often available only within a dialect of an existing language and so face the classic ``chicken-and-egg'' problem: a language like SugarJ \cite{erdweg2011sugarj} or Wyvern \cite{TSLs} (which permit only syntax extensions over a fixed semantics) must overcome the same extrinsic barriers as a language like HJ.
-Second, giving abstraction providers too much control over the language can permit the definition of conflicting  extensions, a problem that is detected ``too late'': when extensions are combined by abstraction clients, who are not necessarily  capable of understanding and fixing the conflict.% Too many constraints can leave it difficult to express real-world abstractions.%It is difficult to guarantee that extensions are sound in isolation and more so to guarantee that they can be {safely} combined.
+Unfortunately, the mechanisms that are available today have several problems.
+First, they do not generally preclude the possibility of  syntactic and/or semantic conflicts between separately developed constructs.
+These conflicts, if detected at all, are detected far too late: when a client happens to use the conflicting constructs within the same program.
+The client must then patch the problem on their own, once again violating the principle of modularity.
+The few exceptions to this are mechanisms that support the definition of mutually orthogonal  syntactic sugar, as in Copper \cite{conf/pldi/SchwerdfegerW09} and our recent work on Wyvern \cite{TSLs,sac15}, but these are built atop a fixed semantics.
+Second, existing mechanisms, including those just mentioned, are often themselves  tied to an obscure dialect or compiler infrastructure and so face the classic ``chicken-and-egg'' problem: they must overcome the same steep barriers to adoption as a dialect like HJ.
+%Clients are not generally capable of understanding and fixing the issue productively.% Too many constraints can leave it difficult to express real-world abstractions.%It is difficult to guarantee that extensions are sound in isolation and more so to guarantee that they can be {safely} combined.
 %This requires further research before it would be appropriate to widely rely on them (which we will discuss as we go on).
@@
 The syntax of each fragment can be characterized by its contributions to the \emph{concrete syntax} (e.g.
-expression forms \lstinlinep{(e, e)}, \lstinlinep{e.l} and \lstinlinep{e.r} and type form \lstinlinep{T * T}) and the \emph{abstract syntax}, where terms are written uniformly by applying a \emph{term constructor} to an \emph{index} and zero or more \emph{arguments}, e.g.
+expression forms \lip{(e, e)}, \lip{e.l} and \lip{e.r} and type form \lip{T * T}) and the \emph{abstract syntax}, where terms are written uniformly by applying a \emph{term constructor} to an \emph{index} and zero or more \emph{arguments}, e.g.
 $\FF{pair}[\FF{triv}](e_1; e_2)$ or $\FF{prj}[\FF{left}](e)$.
@@
 
-Our focus in this paper is to enable the introduction of new static type system fragments as safely composable libraries for an existing and widely-used language, leaving its concrete and abstract syntax fixed.
-This choice is freeing: questions of syntactic ambiguity no longer need to be addressed.
-Moreover, many tools operate by exhaustive case analysis over the concrete or abstract syntax of the terms of the language they work with (e.g.
-syntax highlighters, editor modes, pretty-printers, style checkers, documentation generators).
-These tools will continue to work.
+Our aim in this paper is to take an initial step toward addressing both of these problems by 1) introducing a mechanism that allows library providers to define new type system fragments modularly, and, perhaps most importantly, 2) implementing this mechanism itself as a library, \texttt{typy}, for a widely adopted language, Python.
+%Programmers can construct types and, using these, define typed functions, from within a Python script.
+Typed functions, unlike regular Python functions, go through a process of static typechecking and translation.
+%This process can occur either ahead-of-time or just-in-time.
+
+This design constrains us: we must remain within the confines of Python's existing syntax.
+But in fact, we find that this constraint is freeing: questions about syntactic ambiguity never arise, the syntax is familiar and tools that require an exhaustive syntax specification (e.g.
+syntax highlighters, editor plugins, linters and documentation generators)  work without modification.
+In other words, this design largely sidesteps the \emph{expression problem} \cite{wadler1998expression}.%Taken together, this design addresses many more of the the extrinsic concerns about adoption than a dialect-oriented approach.
 %Indeed, enabling the decentralized introduction of new term constructors is the standard example used when describing the more general problem of enabling the extension of a sum type with new cases as well as new  functions over its cases  \cite{Reynolds75}, prompting Wadler to dub it the \emph{expression problem} \cite{wadler1998expression}.
 
-\paragraph{Approach} Our key insight is that few tools operate by exhaustive case analysis over type constructors.
-We introduce a mechanism that allows library providers to introduce new type system fragments by defining only new type constructors, in libraries, while leaving the concrete and abstract syntax, and the base semantics of the language, fixed.
-
-To support new operations, the semantics of terms are no longer determined directly based on the identity of their term constructors (the usual ``syntax-directed'' approach).
+We show that you do not need an extensible syntax to support an extensible statically-typed  semantics.
+The trick is 1) to organize the semantics around a protocol that delegates control over typechecking and translation of each term to logic associated with some contextually-relevant \emph{type constructor} (\emph{tycon}, for brevity); and 2) to give the ability to define new tycons and the logic associated with them to library providers.
+We call this protocol \textbf{active typechecking and translation} (AT+T).
+
+For example, using \lip{typy}, we will show how a library provider can introduce record types, like those that are built in to languages like ML and Haskell, by defining a tycon called \lip{record}.
+A different library provider, inspired to give a static semantics to a Javascript-style prototypic object system, can define the active tycon \lip{proto}.
+We will discuss several other examples throughout the paper.
+
+Types are constructed by applying a tycon to a \emph{type index} using the subscript operator: \lip{tycon[idx]}.
+For example, we can construct a record type by applying \lip{record} to a mapping from row labels, represented as strings, to types:
+\begin{lstlisting}[numbers=none]
+R = record[
+  'label1' : dyn,    # we discuss dyn and string later
+  'label2' : string
+]
+\end{lstlisting}
+Because the language of types, type indices and tycons is Python itself, we are able to use \emph{array slices} to emulate conventional syntax for row specifications, and assignment to define a type synonym, \lip{R}.% We assume that the types \lip{dyn} and \lip{string}, which we will return to shortly, were imported previously.
+
+The following prototypic object type has an analagous field structure (and for simplicity, the default prototype):
+\begin{lstlisting}[numbers=none]
+P = proto[
+  'label1' : dyn,   
+  'label2' : string 
+]
+\end{lstlisting}
+
+The first question that one should ask when shown a new type is ``how do I introduce a value of this type?'' Values of \lip{typy} types are introduced inside \emph{typed functions}, which are those placed under the control of \lip{typy} by the placement of a decorator.
+Putting the details of typed functions aside for a moment, the form in Python's syntax most similar to the conventional introductory form for records in ML/Haskell is the dictionary literal form, so we would like to be able to introduce values of type \lip{R} inside a typed function using syntax like this:
+\begin{lstlisting}[numbers=none]
+{'label1': val1, 'label2': val2} 
+\end{lstlisting}
+But this is also the form that one would hope to use to introduce a value of the prototypic object type \lip{P} above.
+Moreover, we still need to be able to construct Python dictionaries, of type \lip{dyn}.
+
+Inspired directly by our recent work on \emph{type-specific languages} in Wyvern \cite{TSLs}, we resolve this sort of ambiguity using \emph{local type inference} \cite{Pierce:2000:LTI:345099.345100}.
+When AT+T encounters an expression  of a literal form, it delegates semantic control over it to the tycon of the type that it is being analyzed against.
+For example, when the expression above appears as an argument to a function \lip{f}, then the argument type of \lip{f} determines its meaning:
+\begin{lstlisting}[numbers=none]
+f({'label1': val1, 'label2': val2})
+\end{lstlisting}
+ In positions where there is no expected type, like at the top level of a typed function, the programmer must provide an explicit type ascription.
+We co-opt Python's array slice syntax for type ascriptions:
+\begin{lstlisting}[numbers=none]
+myRecord = {'label1': val1, 'label2': val2} [: R]
+\end{lstlisting}
+Ascriptions can be placed at binding sites as well:
+\begin{lstlisting}[numbers=none]
+myObj [: P] = {'label1': val1, 'label2': val2}
+\end{lstlisting}
+
+The second question one should ask when presented with a new type is ``what can I do with a value of this type?'' For records, the fundamental operation is row projection.
+Similarly, for prototypic objects, it is field projection.
+In both cases, the most natural form in Python's syntax for projecting out a row/field labeled \lip{label1} is \lip{e.label1}, where \lip{e} is the record/object, so we must again come up with some way to resolve the ambiguity.
+Once again, our solution is to take a {tycon-directed} approach: AT+T delegates static control to  the tycon of the type of \lip{e}.
+%Symmetric binary forms, e.g.
+addition, have only a slightly more involved delegation protocol.
+
+We will see exactly how  tycons like \lip{record} and \lip{proto} statically control typechecking and translation (to standard Python code) of expressions that they are delegated control over in Sec.
+\ref{usage}.
+For now, it suffices to state that the code involved is essentially identical to what one would write if asked to implement a compiler for a dialect that built in the construct in question directly.
+The novelty of this paper comes not in the constructs themselves, but in how AT+T decentralizes control over the semantics.
+The locus of control is no longer the language definition but instead individual type constructors, defined in libraries.
+
+% A statically-valued function 
+% %, identified based on the name Python's grammar gives to the form, here \lip{syn_Attribute},  %So if \lip{e} is of record type, then the \lip{record} tycon  is delegated control, and if \lip{e} is of prototypic object type then \lip{proto} is delegated control, in either case by calling \lip{syn_Attribute} with a reference to the syntax tree of the term as well as a \emph{typing context}, which we will describe later.
+% must  either synthesize a type for the term as a whole or indicate a type error.
+If it synthesizes a type, the tycon must then generate a \emph{translation} for the term as a whole, which determines its dynamic behavior.
+So, for the term \lip{e.label1}, if \lip{e} is of record type, then the \lip{record} tycon is delegated control.
+It will check that \lip{label1} is one of the row labels specified in the index of the type of \lip{e}.
+Records can be implemented by translation to tuples (the row labels are only needed statically), so the translation it then generates will project out the appropriate position from the translation recursively generated for \lip{e}.
+
+% We will detail how tycons are implemented as Python classes in Sec.
+\ref{usage}.
+The innovation is in how this protocol decentralizes control over a fixed set of forms, allowing us to define a range of semantic constructs in libraries.
+The language itself needs to specify only the \emph{delegation protocol} that determines how a tycon is delegated control over each syntactic form.
+
+% More specifically, what this means is that the statically valued function  because that the tycon defines must either synthesize a type for the term as a whole or indicate a type error.
+If a type is synthesized, the tycon must also generate a \emph{translation} for the term as a whole.
+
+% We call a type constructor equipped with a set of such statically valued functions an \emph{active type constructor}.
+
+% More specifically, the semantics are a \emph{bidirectionally-typed type-directed translation semantics}.
+
+% allow the associated operations, the semantics of terms are no longer determined directly based on the identity of their term constructors (the usual ``syntax-directed'' approach).
 Instead, we delegate to a statically defined function associated with a relevant type constructor.
-The term constructor's identity only serves to determine how the delegate is assigned.
-We call user-defined type constructors equipped with these static functions and distributed in libraries \emph{active type constructors}.
-
-As a simple example, consider the concrete term \lstinlinep{e.r} (where \lstinlinep{e} is an expression and \lstinlinep{r} is a static label), which maps to the abstract term $\FF{attr}[\text{``r''}](e)$ (i.e.
-the term constructor $\FF{attr}$, \emph{indexed} by a static representation of the label and taking one \emph{argument}, $e$).
-A fragment defining statically typed pairs might see this term as referring to projection of the right component of the pair $e$, while a fragment defining functional record types might want it project out a field labeled \verb|r| from $e$, and an object system might introduce a more complex semantics involving dynamic dispatch.
+The term constructor's identity only serves to determine how to determine this delegate.
+For example, consider the concrete term \lip{x.r} (where \lip{x} is a variable and \lip{r} is a static label), which, according to Python's grammar, maps to the abstract term \lip{Attribute("r"; Name("x"))} (i.e.
+the term constructor $\FF{Attribute}$, \emph{indexed} by a static representation of the label and one \emph{argument}, $e$).
+There are many possible meanings for this term.
+If $e$ is a pair, this might be a way to project out its right component.
+If $e$ were a functional record,  this might be the way to project out a field labeled \verb|r|.
+An object system might introduce a more complex semantics involving dynamic dispatch.
 The type of dynamically classified Python values would follow Python's own complex field lookup protocol.
-Were we to naively ``open up'' the syntax and semantics, these fragments would conflict.
-
-Instead, our semantics will delegate responsibility for assigning a type and a translation to this term to a static function associated with the type constructor of the type recursively synthesized by $e$, the \emph{target} of the operation.
+Were we to naively ``open up'' the syntax and semantics in an attempt to allow all of these features into our language, these semantics would conflict.
+
+% Instead, our semantics will delegate responsibility for assigning a type and a translation to this term to a static function associated with the type constructor of the type recursively synthesized by $e$, the \emph{target} of the operation.
 Thus, tuples, records, objects, dynamic values and variants thereof can  share this syntax.
 
-As we will detail, terms for which there is no clear ``target'' from which to determine a delegate (e.g.
-number or list literals) instead either use the type they are being analyzed against (following the approach for composable syntax extensions as supported in Wyvern \cite{TSLs}) or, if not available, a ``default'' delegate determined by the type constructor governing the function in which they appear.
-Symmetric binary terms (e.g.
-addition) use a third protocol, which we will discuss.
-There is always exactly one active type constructor delegated  responsibility over a term and this delegate is stable with respect to additional extension imports, so semantic ambiguities cannot arise by construction (unlike systems that treat the type system as a ``bag of rules'').
-%This is a generalization of a previous mechanism that we have developed, which deals only with extending the elaboration, but not the typechecking, of the bodies of delimited literals \cite{omar-ecoop14}.
+% An expression like \lip{e.label1} has a clear ``target'' subexpression that can be examined recursively to determine the tycon to delegate control to.
+For other forms, it is perhaps less clear.
+For example, the curly-brace-delimited form in the following expression could be a value of record type, or prototypic object type or be a dynamically typed Python dictionary, in which case it would have type \lip{dyn}:
+% \begin{lstlisting}[numbers=none]
+% f({label1: "val1", label2: "val2"})
+% \end{lstlisting}
+
+% For example, if \lip{f} is a function that takes an argument of a  record type, like the one defined above, then \lip{record} is delegated control over the curly-brace-delimited form.
+This can be understood as a variant of the mechanism that Wyvern introduced for resolving ambiguous syntax extensions \cite{TSLs}, here applied in a setting where there is a fixed syntax but an extensible semantics.% When a literal form appears at the top-level of a function, where there is no expected type, the programmer can provide a type annotation on an expression directly, using the form \lip{e [: T]}, where \lip{e} is an expression and \lip{T} is a type (or a tycon applied to an incomplete index, which we discuss in Sec.
+\todo{inc indices}).
+
+% Symmetric binary terms (e.g. addition) use a third protocol, which we will discuss.
+
+Notably, the semantic flexibility of AT+T does not come at the risk of ambiguity  because there is always exactly one tycon delegated  control over each expression and this delegate remains stable no matter which other tycons have been defined.
+This stands in contrast to systems that treat the type system as a ``bag of rules'', e.g.
+Qi/Shen (a statically typed Lisp dialect) \cite{tarver2008functional}.%This is a generalization of a previous mechanism that we have developed, which deals only with extending the elaboration, but not the typechecking, of the bodies of delimited literals \cite{omar-ecoop14}.
 
@@
 
-\paragraph{Specific Aims} Our specific aims in this paper are to introduce active type constructors and demonstrate that they are expressive, practically realizable within existing languages and theoretically well-founded.
+\paragraph{Organization} The remainder of the paper is organized as follows:
+\begin{itemize}
+\item We provide more details on AT+T as realized in \verb|typy| by tracing through a complete program in Sec.
+\ref{usage}, covering \verb|typy|'s notion of {phase separation}, function types, the delegation protocol, the compilation model, interoperability with Python  and various conveniences, including \emph{incomplete ascriptions}.
+\item We outline more advanced examples of type system fragments that can be implemented as libraries using \verb|typy| in Sec.
+\ref{examples}: functional datatypes, as in ML and Haskell, and a typechecked foreign function interface to a low level language, OpenCL.
+\item We give a minimal specification of AT+T using a typed lambda calculus, @$\lambda$, in Sec. 4. The theoretically minded reader may skip ahead to this section first.
+\item We compare active typing to various related work in Sec. 5.
+\item We conclude after discussing some limitations and directions for future work in Sec. 6. We also discuss the features necessary to implement active typing within other languages.
+\end{itemize}
+
+% Our specific aims in this paper are to introduce this approach, which we call \emph{active typechecking and translation}, and demonstrate that they are expressive, practically realizable within existing languages and theoretically well-founded.
 Our target audience is typed language fragment providers (researchers or domain experts) and language designers interested in extensibility mechanisms.
@@
 normal developers) in writing and reasoning about a variety of real-world programs (this first requires adoption by our target audience).
-We describe two artifacts in pursuit of these aims:
-\begin{enumerate}
-\item We build our mechanism as a library inside Python, \verb|atlang|,  using its quasiquotation and reflection facilities.\footnote{The name was chosen as an initialism for ``actively  typed language'' and because Python marks function decorators, which we use extensively for this purpose, using the @ symbol.} Python is also an interesting starting point  because it  can be thought of as beginning with a  static type system providing just one trivially indexed type constructor, \texttt{dyn} \cite{pfpl}.
+% We describe two artifacts in pursuit of these aims:
+% \begin{enumerate}
+% \item We build our mechanism as a library inside Python, \verb|typy|,  using its quasiquotation and reflection facilities.\footnote{The name was chosen as an initialism for ``actively  typed language'' and because Python marks function decorators, which we use extensively for this purpose, using the @ symbol.} Python is also an interesting starting point  because it  can be thought of as beginning with a  static type system providing just one trivially indexed type constructor, \texttt{dyn} \cite{pfpl}.
 Thus, our mechanism represents a modularly extensible gradual type system.
-Taken together, @\texttt{lang} is almost completely backwards compatible with the Python ecosystem with remarkably little effort: we inherit a large number of  existing tools for working with Python files, Python's substantial packaging infrastructure and access to all existing Python libraries.
-We introduce @\texttt{lang}  in Sec.
+Taken together, \texttt{typy} is almost completely backwards compatible with the Python ecosystem with remarkably little effort: we inherit a large number of  existing tools for working with Python files, Python's substantial packaging infrastructure and access to all existing Python libraries.
+We introduce \texttt{typy}  in Sec.
 \ref{usage}.
@@
 \ref{examples} by briefly describing  some more sophisticated examples, including statically typed functional datatypes with nested pattern matching and a complete and statically safe FFI to OpenCL for GPU programming.
-\item In Sec.
+% \item In Sec.
 \ref{theory}, we describe active type constructors in their essential form, developing a minimal calculus called @$\lambda$.
@@
 For more theoretically minded readers or those with no familiarity with Python, this section can be read first.
-\end{enumerate}
-%@\texttt{lang} can be used from the shell, producing source files that can be further compiled or executed by external means.
-@\texttt{lang} functions targeting a language with Python bindings (e.g.
+% \end{enumerate}
+%\texttt{typy} can be used from the shell, producing source files that can be further compiled or executed by external means.
+\texttt{typy} functions targeting a language with Python bindings (e.g.
 Python itself, as well as the others just described) can also be compiled and invoked interactively from Python.
-%@\texttt{lang} supports a form of staged \emph{ad hoc} polymorphism (based on singleton types) wherein the ``dynamic type'' of a Python value can determine a static type and trigger active typechecking and translation ``just-in-time'' when Python code calls a generic @\texttt{lang} function.
+%\texttt{typy} supports a form of staged \emph{ad hoc} polymorphism (based on singleton types) wherein the ``dynamic type'' of a Python value can determine a static type and trigger active typechecking and translation ``just-in-time'' when Python code calls a generic \texttt{typy} function.
 We show how this can significantly streamline interactive workflows (e.g.
@@
 plotting) but call into a statically-typed  language via a foreign-function interface (FFI) for performance- and safety-critical portions.
-We provide a complete implementation of the OpenCL type system (a language for low-level data parallel programming on GPUs), show several higher-level extensions to it, and discuss bindings that bridge between the type systems of @\texttt{lang}, OpenCL and Python's standard numerics package, \lstinlinep{numpy}, all as a library (contribution 3, Sec.
+We provide a complete implementation of the OpenCL type system (a language for low-level data parallel programming on GPUs), show several higher-level extensions to it, and discuss bindings that bridge between the type systems of \texttt{typy}, OpenCL and Python's standard numerics package, \lip{numpy}, all as a library (contribution 3, Sec.
 \ref{targets}).%We will discuss in detail as our primary realistic case study.
@@
 %The remainder of the paper is organized as follows: in Sec.
-\ref{usage}, we describe the basic structure and usage of @\texttt{lang} with an example that uses several interesting @\texttt{lang} type libraries.
+\ref{usage}, we describe the basic structure and usage of \texttt{typy} with an example that uses several interesting \texttt{typy} type libraries.
 In Sec.
@@
 In Sec.
-\ref{theory}, we reduce this mechanism to  a core lambda calculus, $\lam@\texttt{lang}$,  clarify the relationship to   type-level computation and show how type safety can be maintained.
+\ref{theory}, we reduce this mechanism to  a core lambda calculus, $\lam\texttt{typy}$,  clarify the relationship to   type-level computation and show how type safety can be maintained.
 In Sec.
@@
 
-In Sec.
+%In Sec.
 \ref{related}, we describe how active type constructors relate to several threads of related work.
@@
 
-\section{Active Type Constructors in @\texttt{lang}}\label{usage}
+\section{\texttt{typy}}\label{usage}
+
+%
+Listing \ref{example} shows an example of a well-typed \texttt{typy} program. We will refer to it throughout this section.
+
+\paragraph{Phase Separation} The top level of a \texttt{typy} program is simply a Python script (versions 2.6+ and 3.0+ are supported).
+The execution of this script begins the \emph{static phase} of compilation, as suggested by line \ref{hello-ct}.
+The main purpose of this phase is to define the tycons and types that will be used by the \emph{typed functions} in the program, marked in this example by the decorator \lip{@fn}.
+The static phase ends once  these typed functions have gone through the process of active typechecking and translation.
+The generated translation is what ultimately determines the run-time behavior of the program.
+% The evaluation of the translation of a typed function represents the \emph{dynamic phase}.
+%For example, the string on line \ref{hello-ct} is a statically valued string.
+Types are also statically valued, as we will discuss shortly.
+
+%Code inside , is evaluated during the dynamic phase, i.e.
+these functions can only be called after the active typechecking and translation process we will describe below is complete.
+%There are two ways to transition from the static phase to the dynamic phase, \emph{standalone compilation}, which we will discuss in Sec.
+\ref{external-compilation} and Sec.
+\
+% For example, the string on line \ref{account-start}, discussed further below, is \emph{dynamically valued}.
+%Of particular note, types are statically valued.%No features specific to its primary implementation, CPython, are needed (so \texttt{typy} should support alternative implementations like Jython and PyPy, though as of this writing this claim has not been tested).
+
+\paragraph{Package Management} Using Python as \verb|typy|'s static language pays off immediately on the first line: \texttt{typy} uses Python's import mechanism directly, so Python's package management tools (e.g.
+\lip{pip}) and package repostories (e.g.
+\lip{PyPI}) are directly available for distributing \texttt{typy} libraries, including those defining type system fragments.
+In this case, all of the constructs we will use are part of the \verb|typy| standard library, \texttt{typy.std}, though \texttt{typy}'s semantics do not fundamentally privilege it.
+All existing Python libraries can be used from typed functions, as we will discuss later.% Here, \lip{typy.std} is the ``standard library'' but does not benefit from special support in \lip{typy} itself.% Nothing in \lip{typy.std} benefits from special support from \lip{typy}, the core language package.
+We will see \lip{typy} being used directly below.
+%\texttt{typy} can be installed with a single command: \emph{omitted for review}.
+%\texttt{typy} is agnostic about the particular design choices that we made in our examples.
+
+%The top-level statements in an \texttt{typy} file, like the \lip{print} statement on line 10, are executed to control the compile-time behavior, rather than the run-time behavior, of the program.
+%That is, Python serves as the \emph{compile-time metalanguage} (and, as we will see shortly, the \emph{type-level language}) of \texttt{typy}.
+%For readers familiar with C/C++, Python can be thought of as serving a role similar to (but more general than) its preprocessor and template system (as we will see).
+%\subsection{Active Typechecking}
+
+% Python is dynamically typed, but functions governed by \verb|typy| (those decorated with \lip{@fn} in this example) will be statically typed (Section \ref{s:atc}) and then translated to dynamically typed Python code (Section \ref{s:atr}), both under the control of the type constructors of the types used within them.
+These steps can occur either during a standalone compilation phase (Section \ref{compilation}) or just-in-time upon their first invocation (Section \ref{interactive}).
+
+\paragraph{Types}
 
@@
 \begin{lstlisting}
-^\label{imports-start}^from atlib import record, decimal, dyn, string, 
-  string_in, proto, fn, printf^\label{imports-end}^
-
-^\label{hello-ct}^print "Hello, compile-time world!"
-
-^\label{decl-Details}^Details = record[
+^\label{imports-start}^from typy.std import (record, string, string_in, 
+  proto, decimal, dyn, fn, printf) ^\label{imports-end}^
+
+^\label{hello-ct}^print "Hello, static world!"
+
+^\label{decl-Account}^Account = record[ 
+  'name' : string, 
+  'account_num' : string_in[r'\d{2}-\d{8}']
+]
+
+^\label{decl-Transfer-start}^Transfer = proto[
   'amount' : decimal[2], 
-  'memo' : dyn]
-^\label{decl-Account-start}^Account = record[
-  'name' : string, 
-  'account_num' : string_in[r'\d{2}-\d{8}']]^\label{decl-Account-end}^
-^\label{decl-Transfer}^Transfer = proto[Details, Account]
-
-^\label{log-transfer-start}^@fn[Transfer, ...]
+  'memo' : dyn,
+  'account' :: Account # this field is the prototype
+]^\label{decl-Transfer-end}^
+
+^\label{log-transfer-start}^@fn
 def log_transfer(t):
-  """Logs a transfer to the console."""
-  printf("Transferring %s to %s.", 
+  ^\label{log-transfer-docstring}^"""Logs a transfer to the console."""
+  ^\label{log-transfer-sig}^{t : Transfer}
+  ^\label{log-transfer-print}^printf("Transferring %s to %s.", 
     t.amount.to_string, t.name)^\label{log-transfer-end}^
 
-^\label{main-start}^@fn[...]
+^\label{main-start}^@fn
 def __main__():
-  ^\label{account-start}^account = {name: "Annie Ace", 
-    account_num: "00-00000001"} [:Account]^\label{account-end}^
-  log_transfer(({amount: 5.50, memo: None}, account))
-  log_transfer(({amount: 15.00, memo: "Rent payment"}, 
-    account))^\label{main-end}^
-
-print "Goodbye, compile-time world!"
+  ^\label{account-start}^account = {
+    'name': "Annie Ace", 
+    'account_num': "00-00000001"} [: Account]^\label{account-end}^
+  ^\label{log_transfer-start}^log_transfer({'amount': 5.50, 'memo': None, 
+    'account': account})
+  log_transfer({'amount': 15.00, 'memo': "Rent",
+    'account': account})^\label{main-end}^
+
+print "Goodbye, static world!"
 \end{lstlisting}
-\caption{[\texttt{listing\ref{example}.py}] An @\texttt{lang} compilation script.}
+\caption{[\texttt{listing\ref{example}.py}] A \texttt{typy} program.}
 \label{example}
@@
 
-%
-Listing \ref{example} shows an example of a well-typed @\texttt{lang} program for which stronger correctness guarantees than Python can statically provide have been established using several type system fragments defined in libraries.
-As suggested by line \ref{hello-ct}, the top level of every @\texttt{lang} file can be seen as a \emph{compilation script} written directly in Python.
-@\texttt{lang} requires no modifications to the language (version 2.6+ or 3.0+) or features specific to its main implementation, CPython (so @\texttt{lang} supports alternative implementations like Jython and PyPy).
-This choice pays off immediately on the first line: @\texttt{lang}'s import mechanism is Python's import mechanism, so Python's build tools (e.g.
-\lstinlinep{pip}) and package repostories (e.g.
-\lstinlinep{PyPI}) are directly available for distributing @\texttt{lang} libraries defining type system fragments.
-Here, \lstinlinep{atlib} is the ``standard library'' but does not benefit from special support in \lstinlinep{atlang} itself.% Nothing in \lstinlinep{atlib} benefits from special support from \lstinlinep{atlang}, the core language package.
-We will see \lstinlinep{atlang} being used directly below.
-%@\texttt{lang} can be installed with a single command: \emph{omitted for review}.
-%@\texttt{lang} is agnostic about the particular design choices that we made in our examples.
-
-%The top-level statements in an @\texttt{lang} file, like the \lstinlinep{print} statement on line 10, are executed to control the compile-time behavior, rather than the run-time behavior, of the program.
-%That is, Python serves as the \emph{compile-time metalanguage} (and, as we will see shortly, the \emph{type-level language}) of @\texttt{lang}.
-%For readers familiar with C/C++, Python can be thought of as serving a role similar to (but more general than) its preprocessor and template system (as we will see).
-%\subsection{Active Typechecking}
-
-The language that the compilation script is written in, Python, is dynamically typed, but functions governed by \verb|atlang| (those annotated with \lstinlinep{@fn} in this example) will be statically typed (Section \ref{s:atc}) and then translated to dynamically typed Python code (Section \ref{s:atr}), both under the control of the type constructors of the types used within them.
-These steps can occur either during a standalone compilation phase (Section \ref{compilation}) or just-in-time upon their first invocation (Section \ref{interactive}).
-
-\paragraph{Types}
-Types are constructed programmatically in the compilation script, always taking the form of a type constructor applied to a type index:  \lstinlinep{tycon[idx]}.
-Unlike many statically-typed languages, types are constructed programmatically, rather than  declared ``literally'' (e.g.
+Types are constructed programmatically in the static phase by applying a type constructor to a type index:  \lip{tycon[idx]}.
+%Unlike many statically-typed languages, types are constructed programmatically, rather than  declared ``literally'' (e.g.
 classes in Java, datatypes in ML).
 %Put another way,
-%@\texttt{lang} supports \emph{static computation at the level of types} and Python is its type-level language.
-The static language subsumes th%The compilation script can assign shorter names to types for convenience (removing the need for  facilities like \lstinlinep{typedef} in C or \lstinlinep{type} in Haskell).
-In this example, we see several useful types being constructed:
+%\texttt{typy} supports \emph{static computation at the level of types} and Python is its type-level language.
+The static language subsumes th%The compilation script can assign shorter names to types for convenience (removing the need for  facilities like \lip{typedef} in C or \lip{type} in Haskell).
+In Listing \ref{example}, we construct several types and define short synonyms for them using the standard Python assignment construct:
 \begin{enumerate}
-\item On line \ref{decl-Details}, we construct a functional record type with two (immutable) fields, naming it (by assigning it to a static identifier) \lstinlinep{Details}.
-The field \lstinlinep{amount} has type \lstinlinep{decimal[2]}, which classifies decimal numbers having two decimal places, and \lstinlinep{memo} has type \lstinlinep{dyn}, which comes already constructed, classifying dynamic Python values.
-%Here, \lstinlinep{record} and \lstinlinep{decimal} are {indexed type constructors}.
-For convenience, the index for a \lstinlinep{record} type can be provided using borrowed Python  syntax for array slices (e.g.
-\lstinlinep{a[start:end]}) to approximate conventional notation for type annotations.
-The field names are simply static strings (and could be computed rather than written literally, e.g.
-\lstinline{'a' + 'mount'}, to support features like \emph{type providers} \cite{Syme:2013:TIF:2429376.2429378}).
-\item On line \ref{decl-Account-start}, we construct another record type.
-The field \lstinlinep{name} has type \lstinlinep{string} and the field \lstinlinep{account_num} has a type classifying strings guaranteed statically to be in the regular language specified  by the  regular expression pattern provided as the index (written here as a \emph{raw string literal} \cite{python}).
-This  fragment, based on recent work, statically tracks the regular language an immutable string is a member of through operations like concatenation and substitution, which is useful for ensuring well-formedness and preventing injection attacks \cite{sanitation-psp14}.%To our knowledge, this type system is not  expressible in other simply typed languages (we do not consider dependently-typed logics without a distinct phase separation).
-\item 
-%Prototypic inheritance  is a feature of some dynamically-typed languages (most notably, Self \cite{Ungar:Smith:oopsla:1987} and Javascript). 
-On line \ref{decl-Transfer}, we construct a simple \emph{prototypic object type} \cite{Lie86}, \lstinlinep{Transfer}, classifying values consisting of a \emph{fore} of type \lstinlinep{Details} and a \emph{prototype} of type \lstinlinep{Account}.
-%\footnote{We encourage the convention of naming types that are semantic in camelcase and types and type constructors that are structural in nature, like \texttt{dyn} or \texttt{record}, in lowercase.}
- If a field cannot be found in the fore, the type system will delegate (statically) to the type of the prototype.
-This makes it easy to share the values of the fields of a common prototype amongst many fores, here to allow us to describe multiple transfers to the same account.% differing only in the transaction  details.
+\item On line \ref{decl-Account}, we construct a record type, as discussed in Sec.
+1, and define the synonym \lip{Account}.
+The row labels are written as statically valued strings (and could be statically computed rather than written literally, e.g.
+\lip{'na' + 'me'}).
+\item The row labeled \lip{'account_num'} has the \emph{regular string type} \lip{string_in[r'\\d\{2\}-\\d\{8\}']}.
+Values of a regular string type are  guaranteed to be in the regular language specified by the type index, a regular expression written as a \emph{raw string literal} (per Python's usual conventions).
+In this case, this expresses the invariant that the account number must consist of two digits followed by a dash followed by 8 digits.
+The semantics of this fragment are based on a core calculus we have recently specified in a workshop paper \cite{sanitation-psp14}.
+ %Here, \lip{record} and \lip{decimal} are {indexed type constructors}. 
+%For convenience, the index for a \lip{record} type can be provided using borrowed Python  syntax for array slices (e.g. \lip{a[start:end]}) to approximate conventional notation for type annotations.
+
+\item On line \ref{decl-Transfer-start}, we construct a prototypic object type \cite{Lie86}.
+The field \lip{'amount'} has type \lip{decimal[2]}, which classifies decimal numbers having two decimal places, and the field \lip{'memo'} has type \lip{dyn}, which classifies all dynamically typed Python values.\footnote{The phrase ``dynamically typed'' is widely regarded as as a misnomer from a type theoretic perspective; ``dynamically tagged'' would be more accurate.
+We are guests inside Python, however, so we will not wage this battle here.} The field labeled \lip{'account'} has type \lip{Account}.
+It specifies this with a double colon, which \lip{proto} takes to mean that the value of this field should be used as the prototype.
+This will affect the semantics of the field projection operation: when a field is not present in the object itself, it will defer to the prototype.
+This allows us to  share one account value across multiple transfers, while conveniently allowing clients to access its rows transparently, as if they were fields in the object itself.
+This can be seen as a statically typed variant of Javascript's object system (which is a simplified variant of Self's object system \cite{Ungar:Smith:oopsla:1987}).
+%To our knowledge, this type system is not  expressible in other simply typed languages (we do not consider dependently-typed logics without a distinct phase separation).
 \end{enumerate}
 
-\paragraph{Active Type Constructors} Type constructors are Python classes inheriting from \lstinlinep{atlang.Type} (classes here serving as Python's form of open sums).
-We show portions of \lstinlinep{atlib.fn} in Listings \ref{fn-top} and \ref{fn-stmts}, \lstinlinep{atlib.record} in Listing \ref{record} and \lstinlinep{atlib.string_in} in Listing \ref{string_in}, which we will discuss as we go on.
-The types in our example are instances of these classes.
-
-\paragraph{Incomplete Types} Incomplete types arise from the application of a type constructor to an index containing an ellipsis (a valid array slice form in Python).
-The static terms \lstinlinep{fn[Transfer, ...]} and \lstinlinep{fn[...]} seen on lines \ref{log-transfer-start} and \ref{main-start} are incomplete types.
-We will discuss how the elided portions of an index (e.g.
-the return types, which need not be provided) will be inferred below.
-
-Incomplete types are not instances of \lstinlinep{atlang.Type} but rather instances of \lstinlinep{atlang.IncompleteType}.
-An incomplete type is defined by a type constructor (here \lstinlinep{fn}) and an incomplete index (here, \lstinlinep{(Transfer, Ellipsis)}).
-The language controls whether the form \lstinlinep{tycon[idx]} results in a complte or incomplete type by overloading the subscript operator on the \lstinlinep{atlang.Type} class object (using Python's \emph{metaclass} mechanism \cite{python}).
-
-\paragraph{Function Definitions}
+\paragraph{Type Constructors} Type constructors are classes inheriting from \lip{typy.Type}.
+For example, a portion of \lip{typy.std.record} is shown in Listing \ref{record_init}.
+During the static phase, types are instances of these classes.
+The usual Python instantiation protocol (involving a method named \lip{__init__}) is not used, for reasons that we will cover below.
+Instead, as just described, types are instantiated using the subscript operator.
+This is implemented internally using Python's metaclass mechanism to overload the subscript operator on the class object \cite{python}.
+The tycon can validate and normalize  type indices by defining a class method called \lip{initialize_idx}, as shown in Listing \ref{record_init}.
+%When \verb|typy| delegates control to a tycon during AT+T, discussed below, it will do so by invoking other methods.
+Both \emph{class methods} (i.e.
+methods where the first argument is the tycon itself) and \emph{instance methods} (i.e.
+methods where the first argument is a type constructed by the tycon) are used by \verb|typy|.% It is important to keep in mind that these method calls occur during the static phase.%, and thus AT+T constitutes a form of compile-time metaprogramming.%These methods are, as a point of emphasis, written in Python and invoked statically, by the protocols we will detail in the remainder of this section.
+It is useful in some rhetorical contexts to refer to a tycon equipped with such statically evaluated functions as an \emph{active type constructor}.
+
+\paragraph{Typed Functions}
+
+\begin{codelisting}
+\begin{lstlisting}
+class record(typy.Type):
+  @classmethod
+  def initialize_idx(cls, idx):
+    # turns slices into a dict for convenience
+    idx = _dict_from_slices(idx)
+    # checks that idx is a map from strings to types
+    _validate_sig(idx) # not shown
+    return idx
+
+  # ... more methods defined in later listings ...
+\end{lstlisting}
+\caption{Tycons are classes during the static phase. They have the opportunity to validate and normalize the type index.}
+\label{record_init}
+\end{codelisting}
+
+A \emph{typed function} can be defined by decorating a Python function definition with a tycon, here \lip{fn}, as shown on lines \ref{log-transfer-start} and \ref{main-start} of Listing \ref{example}.
+
+Decorator syntax in Python is syntactic sugar for applying a transformation to the function being decorated \cite{python}.
+Because the decorator value in this case is a tycon, which is a Python class, we must again use Python's  metaclasses and operator overloading, here of the call operator, to enable this usage.
+This is one reason why we must use subscript notation, rather than function application notation, for type construction (another being the usefulness of array slice syntax for record-like types, which is only available inside the subscript form).
+
+The result of this decorator transformation is  an instance of \lip{typy.Function}.
+The underlying Python function is discarded after \lip{typy} extracts its abstract syntax tree and \emph{static environment}, i.e.
+its closure and the globals dictionary of the Python module it was defined within, using Python's reflection capabilities and the \lip{inspect} and \lip{ast} packages in Python's standard library \cite{python}.
+The \lip{ast} package exposes Python's parser and abstract syntax tree.
+The reader is encouraged to refer to its documentation (available online at \url{https://docs.python.org/2/library/ast.html#abstract-grammar}) to fully understand the example code in the remainder of this section.
+
+Like standard Python functions, the body of a typed function can (optionally) begin with a \emph{docstring}.
+Unlike standard Python functions, the line immediately following the docstring specifies a \emph{type signature}, written by repurposing Python's dictionary literal syntax.\footnote{Python 3 introduced syntax for function type ascriptions, but we do not use it because 1) Python 3 is not yet widely adopted, for the reasons discussed in Sec.
+1; and 2) it is awkward when one wishes to introduce parametric polymorphism (an issue we will not discuss at depth in this paper).} For example, the signature on line \ref{log-transfer-sig} of Listing \ref{example} specifies that the argument \lip{t} of \lip{log_transfer} must have type \lip{Transfer}.
+
+The return type of the function is not specified here, so it will be inferred, as we will describe below.
+For clarity and in cases where it cannot be inferred, such as when writing a recursive function, the return type can be specified explicitly by a type signature of the following form:
+ \begin{lstlisting}[numbers=none]
+ {t : Transfer} >> unit
+ \end{lstlisting}
+Typed functions with no arguments and a return type that can be inferred, like \lip{__main__}, can omit the type signature if desired.
+
+As in languages like ML and Haskell, typed functions in \verb|typy| are themselves values of a function type.
+This type is constructed by applying the tycon used as the decorator to a tuple consisting of the argument and return types.
+In this example, \lip{log_transfer} has type \lip{fn[Transfer, unit]}.
+This function type is analagous to function types in ML and Haskell, e.g.
+\lstinline{Transfer -> unit}.\footnote{As of this writing, we have not decided whether to privilege \lip{typy.std.fn} with a monopoly over comparable infix syntax, e.g.
+\lip{Transfer >> unit}.}
+
+\subsection{Active Typechecking}\label{s:atc}
+Let us now trace through the process of typechecking the body of the function \lip{__main__}, which constructs a value of type \lip{Account} and then logs two \lip{Transfer}s by calling \lip{log_transfer}.
+In Sec.
+\ref{s:atr} we cover the subsequent active translation process.
+These two processes together constitute AT+T.
+
+\paragraph{Function Bodies}
+
 \begin{codelisting}[t]
+  % ^\label{normalize_idices_start}^@classmethod
+  % def normalize_idx(cls, idx):
+  %   # ... ensure that idx is a tuple of types ...
+  % @classmethod
+  % def normalize_inc_idx(cls, inc_idx):
+  %   # ... ensure that idx is a tuple of types, 
+  %   #     except return type ...^\label{normalize_idices_end}^
 \begin{lstlisting}
-class fn(atlang.Type):
-  ^\label{fn_init}^def __init__(self, idx): 
-    # called by atlang.Type via the [] operator 
-    atlang.Type.__init__(self, 
-      self._check_and_norm(idx))
-
-  @classmethod
-  ^\label{syn_idx_FunctionDef}^def syn_idx_FunctionDef(cls, ctx, inc_idx, node):
-    arg_types = cls._check_and_norm_inc(inc_idx)
+class fn(typy.Type):
+  ^\label{syn_idx_FunctionDef}^@classmethod
+  def syn_idx_FunctionDef(cls, ctx, inc_idx, node):
     if not hasattr(ctx, 'assn_ctx'): 
-      ctx.assn_ctx = { }
-    ctx.assn_ctx.update(zip(node.args, arg_types))
-    for stmt in node.body: ctx.check(stmt)
-    last_stmt = node.body[-1]
+      ctx.assn_ctx = { } # init. assignables context
+    # add function arguments to assn_ctx
+    ctx.assn_ctx.update(zip(node.args, inc_idx[:-1]))
+    # check each statement in body (post-type sig.)
+    for stmt in node.proper_body: ctx.check(stmt)
+    # synthesize return type from last stmt
+    last_stmt = node.proper_body[-1]
     if isinstance(last_stmt, ast.Expr): 
@@
     else: rty = unit
-    return (arg_types, rty) # complete index
+    # generate complete idx
+    return inc_idx[:-1] + (rty,)
 
   ^\label{ana_FunctionDef}^def ana_FunctionDef(self, ctx, node):
+    arg_tys, rty = self.idx[0:-1], self.idx[-1]
     if not hasattr(ctx, 'assn_ctx'): 
-      ctx.assn_ctx = { } # top-level functions
-    ctx.assn_ctx[node.name] = self # recursion
-    ctx.assn_ctx.update(zip(node.args, self.idx[0]))
-    # all but last
-    for stmt in node.body[0:-1]: ctx.check(stmt)
-    last_stmt = node.body[-1]
+      ctx.assn_ctx = { } # init. assignables context
+    # add fn name (for recursion) + args to assn_ctx
+    ctx.assn_ctx[node.name] = self
+    ctx.assn_ctx.update(zip(node.args, arg_tys))
+    # check all but last statement in body
+    for stmt in node.proper_body[0:-1]: ctx.check(stmt)
+    last_stmt = node.proper_body[-1]
+    # for last statement, if expr, analyze against rty
     if isinstance(last_stmt, ast.Expr): 
-      ctx.ana(last_stmt.value, self.idx[1])
-    elif self.idx[1] == unit or self.idx[1] == dyn: 
+      ctx.ana(last_stmt.value, rty)
+    # otherwise, rty must be unit or dyn
+    elif rty == unit or rty == dyn: 
       ctx.check(last_stmt)
-    else: raise atlang.TypeError("...", last_stmt)
-\end{lstlisting}
-\caption{A portion of the type constructor \texttt{atlib.fn}.}
-\label{fn-top}
-\end{codelisting}
- To be typechecked by @\texttt{lang} and define run-time behavior, function definitions must have an \emph{ascription}, provided by \emph{decorating} them with either a type or, here, an incomplete type, e.g.
-on lines 14 and 20 (see \cite{python} for a discussion of Python decorators; we again use Python's metaclasses and operator overloading to support the use of a class as a decorator).
-
- Ascribed function definitions do not share Python's semantics and indeed the underlying Python function is discarded immediately after its abstract syntax and static environment (its closure and the global dictionary of the Python module it was defined in) have been extracted by the compiler using Python's built-in reflection capabilities and \lstinlinep{inspect} and \lstinlinep{ast} packages \cite{python}.
-The \lstinlinep{ast} package defines Python's term constructors and we will refer to these throughout the paper.
-
-\subsection{Active Typechecking}\label{s:atc}
-We will now trace through how the functions \lstinlinep{log_transfer} and \lstinlinep{__main__} in Listing 1 are typechecked by \verb|atlang|, via delegation to the type constructors just mentioned.
-
-\paragraph{Functions}
-The @$\texttt{lang}$ compiler begins by delegating to the type constructor of the ascription on the function in one of two ways, depending on whether the ascription is complete.
-
-If the ascription is an incomplete type, as in our example, the compiler invokes the class method \lstinlinep{syn_idx_FunctionDef}, seen on line \ref{syn_idx_FunctionDef}, passing it the \emph{compiler context} (an object that provides hooks into the compiler and a storage location for accumulating information during typechecking), the incomplete type index and the syntax tree of the function definition.
-Here, this method: \begin{enumerate}
-\item Checks and normalizes the  incomplete function type index.
-\lstinlinep{_check_and_norm_inc} (not shown) checks that the argument types are indeed types and turns empty and single arguments into a 0- or 1-tuples for uniformity.
-\item Adds the argument names and types to a field of the context that tracks the types of local assignments (initializing it first if it is a top-level function, as in our example).
-\item Asks the compiler, via the method \lstinlinep{ctx.check}, to typecheck each statement in the body (discussed below)
-\item If the term constructor of the last statement is \lstinlinep{ast.Expr} (i.e.
-a top-level expression), then the type of this expression is the return type, otherwise it is \lstinlinep{atlib.unit}.
-Note that this choice of using the last expression as the implicit return value (and considering any statement-level term constructor other than \lstinlinep{Expr} to have a trivial value) is made by \verb|atlib|, not by the language itself.
-\item A complete index is generated, from which a type will be synthesized for the function, here \lstinlinep{fn[Transfer, atlib.unit]} because uses of the \lstinlinep{printf} operator, also defined in \lstinlinep{atlib}, have type \verb|atlib.unit| (i.e.
-\lstinlinep{printf} is used only for its effect).
-\end{enumerate}% A different function type constructor could be written to support \lstinlinep{return} statements (omitted for concision).% The choices is made by clients .
-
-Were the ascription a complete type like \lstinlinep{fn[Transfer, dyn]}, the compiler would delegate to \lstinlinep{fn} by calling an instance method on it, \lstinlinep{ana_FunctionDef}, seen on line \ref{ana_FunctionDef}, rather than a class method.
-This method proceeds similarly, but does not need to perform the first step and last steps (and does not need to be a class method) because the index was already determined when the type was constructed, so it can be accessed via \lstinlinep{self.idx}.
-If the final statement is an expression, it is analyzed against the provided return type (see below), rather than asked to synthesize a type.
-Functions that don't end in expressions can only have return types \lstinlinep{unit} or \lstinlinep{dyn} (again, a choice made by \verb|atlib|).%.
-If it is, the expression .
-% Note that in this example, recursive functions are only supported when the ascription is a type, to avoid complications related to synthesizing a type for a recursive function call.
-%On lines \ref{log-transfer-start}-\ref{log-transfer-end}, we see the function \lstinlinep{log_transfer}.
-We follow Python conventions by starting with a documentation string  that informally specifies the behavior of the function (so standard documentation generators can be used).
-Before moving into the body, however, we  also give an \emph{argument signature} (line \ref{arg-sig}) stating that the type of \lstinlinep{t} is \lstinlinep{Transfer}, the prototypic object type defined above.
-
-%This again repurposes Python's array slice syntax to approximate conventional notation for type annotations.
-% (and following conventions for documentation strings, which also describe ``types'' at the end).
-% The ``keyword'' \lstinlinep{type} was chosen because it is the name of a built-in function in Python (and thus would not be used in the way shown for any other purpose).
-
-\paragraph{Statements}
-\begin{codelisting}[t]
-\begin{lstlisting}
-#class fn(atlang.Type): (cont'd) 
+    else: raise typy.TypeError("<msg>", last_stmt)
+
   @classmethod
   def check_Assign_Name(cls, ctx, stmt):
-    x, e = stmt.target.id, stmt.value # see ast
+    x, e = stmt.target.id, stmt.value # cf. ast docs
     if x in ctx.assn_ctx: ctx.ana(e, ctx.assn_ctx[x])
     else: ctx.assn_ctx[x] = ctx.syn(e)
+
+  ^\label{check_Expr}^@classmethod
+  def check_Expr(cls, ctx, stmt):
+    ctx.syn(stmt.value)
 
@@
       try: return ctx.syn_Lift(ctx.static_env[e.id])
-      except KeyError: raise atlang.TypeError("...", e)
+      except KeyError: raise typy.TypeError("...", e)
 
   ^\label{default_asc}^@classmethod
-  def syn_default_asc_Str(cls, ctx, e): return dyn
+  def syn_default_asc_Str(cls, ctx, e): return string
 \end{lstlisting}
-\caption{Some forms in the body of a function delegate to the type constructor of the function they are defined within (via class methods during typechecking).}
-\label{fn-stmts}
+\caption{A portion of the type constructor \texttt{typy.std.fn}.}
+\label{fn-top}
 \end{codelisting}
-The \lstinlinep{ctx.check} method mentioned above is defined by the compiler.
-As is the pattern, it simply delegates to an active type constructor based on the term constructor of the statement being checked.
-Most statement-level term constructors other than \lstinlinep{Expr} simply delegate to the type constructor of the function they are defined within by calling class methods of the form \lstinlinep{check_TermCon}, where \lstinlinep{TermCon} is a term constructor or combination of term constructors in a few cases where a finer distinction than what Python itself made was necessary.
-
-For example, the class method \lstinlinep{check_Assign_Name}, seen in Listing \ref{fn-stmts}, is called for statements of the form \lstinlinep{name = expr}, as on line \ref{account-start} of Listing \ref{example}.
-In this examples, the assignables context (the \lstinlinep{assn_ctx} field of the compiler context) is consulted to determine whether the name being assigned to already has a type due to a prior assignment, in which case the expression is analyzed against that type using \lstinlinep{ctx.ana}.
-If not, the expression must synthesize a type, using \lstinlinep{ctx.syn}, and a binding is added.
-%This interpretation of the assignment form cannot cause the function it appears in to return so \lstinlinep{None} is returned (Python returns \lstinlinep{None} by default if nothing is explicitly returned) and the subsumption rule applies trivially, so we do not need to define \lstinlinep{ana_rty_Assign_Name} separately.
+After processing the docstring and type signature, if provided, \verb|typy| immediately hands control over the function body to the tycon provided as the decorator, i.e.
+\lip{fn} in this example.
+This happens by one of two methods, depending on whether the return type must be inferred.
+
+If the return type must be inferred, as in Listing \ref{example}, the semantics invokes the class method \lip{syn_idx_FunctionDef}, seen on line \ref{syn_idx_FunctionDef} of Listing \ref{fn-top}, passing it the \emph{semantic context} (an object that provides hooks into the semantics and a storage location for accumulating information during typechecking), the \emph{incomplete type index} (here, the pair \lip{(Transfer, Ellipsis)}, where \lip{Ellipsis} stands in for the unknown return type), and the syntax tree of the function definition.
+This method must return the complete type index (i.e.
+one without an \lip{Ellipsis}).
+The method in Listing \ref{fn-top} proceeds as follows: \begin{enumerate}
+\item It initializes an \emph{assignables context}, which is a dictionary that  tracks assignable variables and their types.
+It is stored in the semantic context.
+Its first entries are the function arguments and their corresponding types, extracted from the incomplete type index (by dropping the last element, written \lip{inc_idx[:-1]}).
+\item It asks \lip{typy}, by invoking the method \lip{ctx.check}, to typecheck each statement in the body (discussed below).
+\item If the abstract form of the final statement is \lip{ast.Expr} (i.e.
+a top-level expression), then its type is used as the return type (typechecking of expressions is also discussed below).
+Otherwise it is \lip{typy.std.unit}.
+Note that this choice of using the last expression as the implicit return value (and considering any statement of a form other than \lip{Expr} to have a trivial value) is made by \verb|fn|, not by \lip{typy}'s semantics.\footnote{The actual implementation of \texttt{fn} supports return statements as well, but we omit this logic here for simplicity.} %In this case, the return type is \lip{unit} because invocations of the \lip{printf} operator, also defined in \lip{typy.std}, have type \lip{unit} (i.e.
+\lip{printf} is used only for its effect).
+We will not cover its details here, but note that the format string can be typechecked statically.
+\item Having determined the return type, the complete type index is  generated and returned (by dropping the \lip{Ellipsis} and concatenating a 1-tuple of the return type).
+\verb|typy| will then use this to construct the function's type as described above.
+\end{enumerate}% A different function type constructor could be written to support \lip{return} statements (omitted for concision).% The choices is made by clients .
+
+Had the type signature specified a return type explicitly, then the type index would not need to be inferred.
+In this case, \lip{typy} constructs the complete type first, i.e.
+\lip{fn[Transfer, unit]}, then invokes its instance method \lip{ana_FunctionDef}, seen on line \ref{ana_FunctionDef} of Listing \ref{fn-tycon}.
+This method proceeds essentially as above, but has access to the complete index as \lip{self.idx}.
+Thus, if the last statement in the function is an expression, it is analyzed against the provided return type, rather than asked to synthesize a type (see below).
+%Functions that don't end in expressions can only have return types \lip{unit} or \lip{dyn} (again, a choice made by \verb|fn|, not the semantics).%.
+If it is, the expression .
+% Note that in this example, recursive functions are only supported when the ascription is a type, to avoid complications related to synthesizing a type for a recursive function call.
+%On lines \ref{log-transfer-start}-\ref{log-transfer-end}, we see the function \lip{log_transfer}.
+We follow Python conventions by starting with a documentation string  that informally specifies the behavior of the function (so standard documentation generators can be used).
+Before moving into the body, however, we  also give an \emph{argument signature} (line \ref{arg-sig}) stating that the type of \lip{t} is \lip{Transfer}, the prototypic object type defined above.
+
+%This again repurposes Python's array slice syntax to approximate conventional notation for type annotations.
+% (and following conventions for documentation strings, which also describe ``types'' at the end).
+% The ``keyword'' \lip{type} was chosen because it is the name of a built-in function in Python (and thus would not be used in the way shown for any other purpose).
+
+\paragraph{Statements}
+% \begin{codelisting}[t]
+% \begin{lstlisting}
+% #class fn(typy.Type): (cont'd) 
+% \end{lstlisting}
+% \caption{Some forms in the body of a function delegate to the type constructor of the function they are defined within (via class methods during typechecking).}
+% \label{fn-stmts}
+% \end{codelisting}
+The \lip{ctx.check} method called on each statement in the function body by the logic in  Listing \ref{fn-top} is provided by \verb|typy|.
+As is the pattern, this method delegates to a tycon determined based on the abstract form of the statement being checked.
+Most statements simply delegate back to the tycon of the function they appear within via a class method named \lip{check_AbsForm}, where \lip{AbsForm} is the name of the statement's abstract form as defined by Python's abstract syntax (or a combination of abstract form names in a few cases where finer distinctions  were necessary).
+
+For example, the class method \lip{check_Assign_Name}, seen in Listing \ref{fn-tycon}, is invoked for statements of the form \lip{name = expr}, as on line \ref{account-start} of Listing \ref{example}.
+In this example, the assignables context (the \lip{assn_ctx} field of the semantic context) is consulted to determine whether the name that is being assigned to already has a type due to a prior assignment, in which case the expression is analyzed against that type using \lip{ctx.ana}.
+If not, the expression must synthesize a type, using \lip{ctx.syn}, and a binding is added to \lip{assn_ctx}.
+%This interpretation of the assignment form cannot cause the function it appears in to return so \lip{None} is returned (Python returns \lip{None} by default if nothing is explicitly returned) and the subsumption rule applies trivially, so we do not need to define \lip{ana_rty_Assign_Name} separately.
+
+The class method \lip{check_Expr}, defined on line \ref{check_Expr} of Listing \ref{fn-tycon}, specifies that top-level expressions must synthesize a type.
+
+There are a number of other statement forms in Python's syntax, but \lip{typy} treats them all similarly, so we omit a detailed discussion of the delegation protocol (it is given in the \lip{typy} documentation).
 
 %Return statements, on the other, require separately defining return type synthesis and analysis logic.
-The class method \lstinlinep{syn_rty_Return} is called for return statements when a return type is not yet known.
-It must synthesize a return type based on the value being returned (defaulting to \lstinlinep{unit}).
-Subsequent return statements are processed by \lstinlinep{ana_rty_Return}, which checks the provided value against the known return type.
-If no value is provided, then the return type must be \lstinlinep{unit}.
-Note that this function does not need to return a value; if a \lstinlinep{atlang.TypeError} is not raised, then analysis succeeds.
-
-\paragraph{Expressions} The methods \lstinlinep{ctx.ana} and \lstinlinep{ctx.syn} are also defined by the compiler and can be called by type constructors to request type analysis (when the expected type is known) and synthesis (when the type is an ``output''), respectively, for an expression.
-Once again, the compiler delegates to a type constructor by a protocol that depends on the term constructor, invoking methods of the form \lstinlinep{ana_TermCon} or \lstinlinep{syn_TermCon}.
-This represents a form of bidirectional type system (sometimes also called a \emph{local type inference system}) \cite{Pierce:2000:LTI:345099.345100}, and the standard subsumption principle applies: if analysis is requested and an \lstinlinep{ana_TermCon} method is not defined but a \lstinlinep{syn_TermCon} method is, then synthesis proceeds and then the synthesized type is checked for equality against the type provided for analysis.
-Type equality is defined by checking that the two type constructors are identical and that their indices are equal.
-Two types with different type constructors are never equal, to ensure that the burden of ensuring that typing respects type equality is local to a single type constructor.
-No form of subtyping or implicit coercion is supported for the purposes of this paper (though we have designed a mechanism that similarly localizes reasoning to a single type constructor, we do not  describe it here.)
-
-\subparagraph{Names} If the term constructor of an expression is \lstinlinep{ast.Name}, as when referring to a bound variable or an assignable location, then the type constructor governing the function the term appears in is delegated to via the class method \lstinlinep{syn_Name} (names, when used as expressions, must synthesize a type).
-We see the definition of this method for \lstinlinep{fn} starting on line \ref{syn_Name} in Listing \ref{fn-stmts}.
-A name synthesizes a type if it is either in the assignables context or if it is in the static environment.
-In the former case, the type that was synthesized when the assignment occurred (by \lstinlinep{syn_Assign_Name}) is returned.
-
-In the latter case, we must lift the static value to run-time and give it an appropriate type.
-This cannot always be done automatically (because it is a form of serialization).
-Instead, any Python value with a \lstinlinep{atlang_syn_Lift} method returning an \lstinlinep{atlang} type can provide explicit support for this operation.
-A few other constructs also have built in support.
-For the purposes of this section, this includes other typed @\texttt{lang} functions (which synthesize their own type), untyped Python functions and classes (which synthesize types \lstinlinep{pyfun} and \lstinlinep{pyclass}, and thus can only be called with values of type \lstinlinep{dyn}), modules (which are given a \emph{singleton type} -- a type indexed by the module reference itself -- from which other values that can be lifted can be accessed as attributes, see below), and immutable Python data structures like numbers, strings and tuples.
-%In the future, we plan to detail how the interface between Python and @\texttt{lang} can be enriched to support moving typed values in and out of untyped Python functions safely (by inserting checks, defined again by a type constructor, at the interface between typed and untyped code), but we do not have enough room to adequately detail this mechanism here (and it is tangential to our specific aims).
-\subparagraph{Literal Expressions and Ascriptions}
+The class method \lip{syn_rty_Return} is called for return statements when a return type is not yet known.
+It must synthesize a return type based on the value being returned (defaulting to \lip{unit}).
+Subsequent return statements are processed by \lip{ana_rty_Return}, which checks the provided value against the known return type.
+If no value is provided, then the return type must be \lip{unit}.
+Note that this function does not need to return a value; if a \lip{typy.TypeError} is not raised, then analysis succeeds.
+%\todo{more examples?}
+
+\paragraph{Expressions} The methods \lip{ctx.ana} and \lip{ctx.syn} are also provided by \verb|typy|.
+As we just saw, \lip{ctx.ana} is invoked by tycons to request type analysis when the expected type of an expression is known.
+The expression and the expected type are provided as arguments.
+When the type of an expression must be locally inferred, \lip{ctx.syn} is invoked, taking only the expression as an argument and returning the synthesized type.
+In both cases, ill-typed expressions raise \lip{typy.TypeError}, which is equipped with an error message and a reference to the location where the error originated.
+
+Like \lip{ctx.check}, these \lip{ctx.ana} and \lip{ctx.syn} methods do very little heavy lifting themselves.
+Instead, they again delegate control to a tycon by invoking a method with a name based on the abstract form of the expression, either \lip{ana_AbsForm} or \lip{syn_AbsForm}.
+We cover how the delegate is chosen for various forms below.
+
+Because we distinguish type analysis from type synthesis, the process of active typechecking can be understood as a novel form of \emph{bidirectional typechecking} \cite{Pierce:2000:LTI:345099.345100}.
+The usual \emph{subsumption principle} applies as follows: if analysis is requested and a \lip{syn_AbsForm} method is defined, then synthesis proceeds and then the synthesized type is checked for equality against the type provided for analysis.
+Type equality is defined by checking that the two type constructors are equal and that their indices are equal, according to Python's usual protocol.
+Two types with different tycons are never equal, to keep the burden of ensuring that typing respects type equality local to a single tycon.%  No form of subtyping is supported for the purposes of this paper.
+
+Let us now discuss how a delegated tycon is chosen for various expression forms in Python's syntax.
+
+\subparagraph{Literal Forms}
+Python's syntax includes literal forms for strings, numbers, tuples, lists, dictionaries and lambda functions.
+As we discussed in Sec.
+1, we would like to be able to use these forms for values of types other than \lip{dyn}.%\footnote{Proponents of static type systems note that a type is usually understood as a  often take issue with the common use of the phrase ``dynamically typed'' for what is actually a system of dynamic \emph{tags}, rather than a purely ``syntactic discipline'' but we are guests within Python so we will abide by its terminology.}
+
+To support this, when \lip{typy} encounters an expression of a literal form  in an analytic position, the tycon of the type it is being analyzed against is delegated control.
+For example, on line \ref{account-start} of Listing \ref{example}, the dictionary literal form appears in an analytic position here because an \emph{explicit type ascription}, written \lip{[: Account]}, was provided.
+Note that this type ascription form repurposes Python's array slice syntax.
+Array slices can still be performed, because this form is only treated as an ascription when it contains a valid type.
+
 \begin{codelisting}[t]
 \begin{lstlisting}
-class record(atlang.Type):
-  def __init__(self, idx): 
-    # Sig is an unordered mapping from fields to types
-    atlang.Type.__init__(self, Sig.from_slices(idx))
+class record(typy.Type): 
+  # ... continued from Listing 2 ...
+
+  ^\label{ana_Dict}^def ana_Dict(self, ctx, e):
+    for lbl, value in zip(e.keys, e.values):
+      if isinstance(lbl, ast.Str):
+        if lbl.s in self.idx.rows: 
+          ctx.ana(value, self.idx[lbl.s])
+        else: raise typy.TypeError("<not found>", f)
+      else: raise typy.TypeError("<not a label>", f)
+    if len(self.idx.fields) != len(e.keys): 
+      raise typy.TypeError("<row missing>", e)
 
   ^\label{syn_idx_Dict}^@classmethod
-  def syn_idx_Dict(self, ctx, partial_idx, e):
-    if partial_idx != Ellipsis: 
-      raise atlang.TypeError("...bad index...", e)
-    idx = []
-    for f, v in zip(e.keys, e.values):
-      if isinstance(f, ast.Name): 
-        idx.append(slice(f.id, ctx.syn(v)))
-      else: raise atlang.TypeError("...", f)
-    return idx
-
-  ^\label{ana_Dict}^def ana_Dict(self, ctx, e):
-    for f, v in zip(e.keys, e.values):
-      if isinstance(f, ast.Name):
-        if f.id in self.idx.fields: 
-          ctx.ana(v, self.idx[f.id])
-        else: raise atlang.TypeError("...", f)
-      else: raise atlang.TypeError("...", f)
-    if len(self.idx.fields) != len(e.keys): 
-      raise atlang.TypeError("...missing field...", e)
+  def syn_idx_Dict(self, ctx, inc_idx, e):
+    if inc_idx != Ellipsis: 
+      raise typy.TypeError("<bad index>", e)
+    return Signature((lbl.s, ctx.syn(value))
+      if isinstance(lbl, ast.Str)
+      else raise typy.TypeError("<bad label>", f)
+      for lbl, value in zip(e.keys, e.values))
 
   ^\label{syn_Attribute}^def syn_Attribute(self, ctx, e):
-    if e.attr in self.idx: return self.idx[e.attr]
-    else: raise atlang.TypeError("...", e)
+    if e.attr in self.idx.labels: 
+      return self.idx[e.attr]
+    else: raise typy.TypeError("...", e)
 \end{lstlisting}
@@
 %      return target.Dyn()
-\caption{A portion of the \texttt{atlib.record} type constructor.}
+\caption{A portion of the \texttt{typy.std.record} type constructor.}
 \label{record}
 \end{codelisting}
-Python designates literal forms for dictionaries, tuples, lists, strings, numbers and lambda functions.
-In @\texttt{lang}, the type constructor delegated control over terms arising from any of these forms is a function of how the typechecker encounters the term.
-
-If the term appears in an analytic position, the type constructor of the type it is being analyzed against is delegated control.
-We see this on line \ref{account-start} of Listing \ref{example}: the dictionary literal form appears in an analytic context, here because an explicit type ascription, \lstinlinep{[:Account]}, was provided.
-Note that the ascription again repurposes Python's array slice syntax (the start is, conveniently,  optional).
-The compiler invokes the \lstinlinep{ana_Dict} method on the type the literal is being analyzed against, which is defined by its type constructor, here \lstinlinep{atlib.record}, shown on line \ref{ana_Dict} of Listing \ref{record}.
-This method analyzes each field value in the literal against the type of the field, extracted from the type index (an unordered mapping from field names to their types, so that type equality is up to reordering).
-The various literal forms inside the outermost form thus do not need an ascription because they appear in positions where the type they will be analyzed against is provided by \lstinlinep{record}.
-For example, the value of the field \lstinlinep{account_num} delegates to \lstinlinep{string_in} via the \lstinlinep{ana_Str} method, shown in Listing \ref{string_in}.
-
-An ascription directly on a literal can also be an incomplete type.
-For example, we can write \lstinline[basicstyle=\ttfamily\small]{[x, y] [:matrix[...]]} or more concisely \lstinline[basicstyle=\ttfamily\small]{[x, y] [:matrix]} instead of \lstinline[basicstyle=\ttfamily\small]{[x, y] [:matrix[i32]]} when we know \lstinline[basicstyle=\ttfamily\small]{x} and \lstinline[basicstyle=\ttfamily\small]{y} synthesize the type \lstinline[basicstyle=\ttfamily\small]{i32}, because the type constructor can use this information to synthesize the appropriate index.
-The decorator \lstinline[basicstyle=\ttfamily\small]{@fn[Transfer, ...]}, discussed previously, can be seen as a partial type ascription on a statement-level function literal.\footnote{In Python 3.0+, syntax for annotating function definitions directly with type-like annotations was introduced, so the entire index can be synthesized.} Lambda expressions support the same ascriptions:
-\vspace{3px}
-\begin{lstlisting}[numbers=none, basicstyle=\ttfamily\scriptsize]
-(lambda x, y: x + y) [:fn[(i32, i32), ...]]
-\end{lstlisting}
-%\vspace{3px}
-\noindent
+
+Because \lip{Account} is a synonym for a record type, \verb|typy| invokes the \lip{ana_Dict} method shown on line \ref{ana_Dict} of Listing \ref{record}.
+This method analyzes each row value in the literal expression against the row type specified in the type index.
+Recall that the type index is an unordered mapping from field names to their types, so that type equality is up to reordering.
+Various error conditions are shown (in practice, the error messages would be more verbose than shown).
+
+The string literal forms inside the outermost record form on line \ref{account-start} of Listing \ref{example} do not need an ascription, because \lip{record} requests analysis against the appropriate row type.
+For example, the value of the field \lip{account_num} delegates to the tycon \lip{string_in} via the \lip{ana_Str} method, shown in Listing \ref{string_in}.
+This method, consistent with the static semantics we have specified in a recent workshop paper, simply statically checks the string against the regular expression provided as the type index \cite{sanitation-psp14}.
+
+Similary, the prototypic object literals passed to \lip{log_transfer} on lines \ref{log_transfer-start}-\ref{main-end} of Listing \ref{example} do not need ascriptions because they will be analyzed by the  argument type of \lip{log_transfer}, i.e.
+\lip{Transfer} (we omit the details of prototypic dispatch for concision; the function application form is discussed below).
+
+\subparagraph{Incomplete Type Ascription} An ascription placed directly on a literal can also be an \emph{incomplete type}.
+Incomplete types arise from the application of a tycon to a type index containing an ellipsis (a valid array slice form in Python; the value of an ellipsis is the constant \lip{Ellipsis}).
+For example, we can write \lstinline[basicstyle=\ttfamily\small]{(e1, e2) [: tuple[...]]}, or equivalently when the index consist solely of an ellipsis, \lstinline[basicstyle=\ttfamily\small]{(e1, e2) [: tuple]}, instead of \lstinline[basicstyle=\ttfamily\small]{(e1, e2) [: tuple[T1, T2]]} when \lstinline[basicstyle=\ttfamily\small]{e1} and \lstinline[basicstyle=\ttfamily\small]{e2} synthesize types \lstinline[basicstyle=\ttfamily\small]{T1} and \lip{T2}, respectively.
+The tycon is asked to synthesize the complete index via a class method named \lip{syn_idx_AbsForm}, where \lip{AbsForm} is the literal form.
+
+Recall that the process for synthesizing the return type of a typed function also involved the use of an \lip{Ellipsis} (perhaps mysterious at the time).
+Indeed, a typed function definition missing a return type in its type signature is treated exactly like a   function literal ascribed an incomplete function type.
+This is more explicit when one considers lambda expression literals, e.g.
+we can write \lip{(lambda x, y: x + y) [: fn[i32, i32, ...]]} when a lambda function is needed in a synthetic position (this is rare in practice; recall that in an analytic position, e.g.
+when calling a higher-order function, no ascription would be needed at all).
+More specifically, the default implementation of the \lip{syn_idx_Lambda} calls \lip{syn_idx_FunctionDef}. For \lip{fn}, this method, shown in Listing \ref{fn-tycon}, was discussed earlier.
+
 Records support partial type ascription as well, e.g.: 
@@
 \begin{lstlisting}[numbers=none,basicstyle=\ttfamily\scriptsize]
-{name: "Deepak Dynamic", age: "27"} [:record]
+{"name": "Deepak Dynamic", "job": "Example"} [: record]
 \end{lstlisting}
-%\vspace{3px}
-\noindent
-The compiler delegates to the type constructor by a class method in these cases (just as we saw above with \lstinlinep{fn}).
-In this case, the class method \lstinlinep{syn_idx_Dict}, shown on line \ref{syn_idx_Dict} of Listing \ref{record}, would be called.
-Because no record index was provided, an index must be synthesized from the literal itself, and thus the values are not analyzed against a type, but must each synthesize a type.
-
-This brings us to the situation where a literal appears in a synthetic position, as the two string literals above do.
-In such a situation, the compiler delegates responsibility to the type constructor of the function that the literal appears in, asking it to provide a ``default ascription'' by calling the class method \lstinlinep{syn_default_asc_Str}, shown on line \ref{default_asc} of Listing \ref{fn-stmts}.
-In this case, we simply return \lstinlinep{dyn}, so the type of the expression above has type \lstinline[basicstyle=\ttfamily\small]{record['name': dyn, 'age' : dyn]}.
-A different function type constructor might make more precise choices.
-Indeed, a benefit of using Python as our static language is that it is relatively straightforward to define a type constructor that allows for different defaults using block-scoped ''pragmas'' in the static language via Python's \lstinlinep{with} statement \cite{python} (details omitted), e.g.
+The implementation of \lip{syn_idx_Dict} is shown on line \ref{syn_idx_Dict} of Listing \ref{record}. It constructs the record signature by requesting type synthesis for each row value.
+
+This, however, puts us in a situation where the string literals appearing as row values above are in a synthetic position without an ascription.
+Here, \lip{typy} delegates control to the tycon of the function that the literal appears in, asking it to provide a ``default ascription'' by calling the class method \lip{syn_default_asc_Str}, shown on line \ref{default_asc} of Listing \ref{fn-tycon}.
+In this case, we simply return \lip{string}, so the type of the expression above is \lstinline[basicstyle=\ttfamily\small]{record['name' : string, 'age' : string]}.
+
+A benefit of using Python as our static language is that it is relatively straightforward to define \lip{fn} instead such that it allows for different defaults using block-scoped ''pragmas'' in the static language via Python's \lip{with} statement \cite{python} (details omitted), e.g.
 \begin{codelisting}[h]
 \begin{lstlisting}[numbers=none]
-with fn.default_asc(Num=i64, Str=string, Dict=record):
-  @fn[...] # : fn[(), record["a": i64, "b": string]]
+with fn.defaults(Num=i64, Str=string, Dict=record):
+  @fn
   def test(): {a: 1, b: "no ascriptions needed!"} 
 \end{lstlisting}
-\caption{Block-scoped settings can be seen by type constructors.}
+\caption{Block-scoped settings for type constructors.}
 \label{defaultasc}
@@
 %\vspace{-3px}
-%This can also be thought of as an incomplete type ascription, which we could write more explicitly \lstinlinep{@fn[...]},  because the literal itself contains the necessary information to synthesize the full index.
+%This can also be thought of as an incomplete type ascription, which we could write more explicitly \lip{@fn[...]},  because the literal itself contains the necessary information to synthesize the full index.
 We show the more universally available notation here because Python 2.\textit{x} is much more widely adopted as of this writing.
@@
 
-\subparagraph{Targeted Expressions} Expression forms having exactly one sub-expression, like \lstinlinep{-e} (term constructor \lstinlinep{UnaryOp_USub}) or \lstinlinep{e.attr} (term constructor \lstinlinep{Attribute}), or where there may be multiple sub-expressions but the left-most one is the only one required, like \lstinlinep{e[slices]} (term constructor  \lstinlinep{Subscript}) or \lstinlinep{e(args)} (term constructor \lstinlinep{Call}), are called \emph{targeted expressions}.
-For these, the compiler first synthesizes a type for the left-most sub-expression, then calls either the \lstinlinep{ana_TermCon} or \lstinlinep{syn_TermCon} methods on that type.
-For example, we see type synthesis for the field projection operation on records defined starting on line \ref{syn_Attribute} of Listing \ref{record}.% We omit the implementation for concision, but \lstinlinep{fn} defines \lstinlinep{syn_Call} as one would expect, by simply checking each argument against the corresponding argument in the index.
+\subparagraph{Lowercase Names} When the form of an expression is \lip{ast.Name} and the identifier begins with a lowercase character, as when referring to the argument \lip{t} of \lip{log_transfer} or the assignable location \lip{account} in \lip{__main__}, the type constructor governing the function the term appears in is delegated control  via the class method \lip{syn_Name}, i.e.
+such names must synthesize a type.
+We see the definition of this method for \lip{fn} starting on line \ref{syn_Name} in Listing \ref{fn-tycon}.
+A name synthesizes a type if it is either in the assignables context or if it is in the static environment.
+In the former case, the type that was synthesized when the assignment occurred (by \lip{syn_Assign_Name}) is returned.
+
+In the latter case, we must lift the static value to run-time and give it an appropriate type.
+This cannot always be done automatically (because it is a form of serialization).
+Instead, any Python value with a \lip{typy_syn_Lift} method returning an \lip{typy} type can provide explicit support for this operation.
+A few other constructs also have built in support.
+For the purposes of this section, this includes other typed \texttt{typy} functions (which synthesize their own type), untyped Python functions and classes (which synthesize types \lip{pyfun} and \lip{pyclass}, and thus can only be called with values of type \lip{dyn}), modules (which are given a \emph{singleton type} -- a type indexed by the module reference itself -- from which other values that can be lifted can be accessed as attributes, see below), and immutable Python data structures like numbers, strings and tuples.
+%In the future, we plan to detail how the interface between Python and \texttt{typy} can be enriched to support moving typed values in and out of untyped Python functions safely (by inserting checks, defined again by a type constructor, at the interface between typed and untyped code), but we do not have enough room to adequately detail this mechanism here (and it is tangential to our specific aims).
+
+\subparagraph{Targeted Expressions} Expression forms having exactly one sub-expression, like \lip{-e} (term constructor \lip{UnaryOp_USub}) or \lip{e.attr} (term constructor \lip{Attribute}), or where there may be multiple sub-expressions but the left-most one is the only one required, like \lip{e[slices]} (term constructor  \lip{Subscript}) or \lip{e(args)} (term constructor \lip{Call}), are called \emph{targeted expressions}.
+For these, the compiler first synthesizes a type for the left-most sub-expression, then calls either the \lip{ana_TermCon} or \lip{syn_TermCon} methods on that type.
+For example, we see type synthesis for the field projection operation on records defined starting on line \ref{syn_Attribute} of Listing \ref{record}.% We omit the implementation for concision, but \lip{fn} defines \lip{syn_Call} as one would expect, by simply checking each argument against the corresponding argument in the index.
 
@@
 \begin{lstlisting}
-class string_in(atlang.Type):
-  def __init__(self, idx):
-    atlang.Type.__init__(self, self._rlang_of_rx(idx))
-
+class string_in(typy.Type):
   def ana_Str(self, ctx, e):
-    _check_rx(self.idx, e.s)
+    _check_rx(e.s, self.idx)
 
@@
 \end{lstlisting}
-\caption{Binary operations in \texttt{atlib.string\_in}.}
+\caption{Binary operations in \texttt{typy.std.string\_in}.}
 \label{string_in}
@@
     if tycon in other_tycon.handles_Add_with:
-      raise atlang.AmbiguousTyconError("...", 
+      raise typy.AmbiguousTyconError("...", 
         tycon, other_tycon) # (other binops analagous)
 \end{lstlisting}
-\caption{For each type constructor definition and binary operator, \texttt{atlang} runs a modular handle set check to preclude ambiguity.}
+\caption{For each type constructor definition and binary operator, \texttt{typy} runs a modular handle set check to preclude ambiguity.}
 \label{check_tycon}
@@
 The major remaining expression forms are the binary operations, e.g.
-\lstinlinep{e + e} or \lstinlinep{e < e}.
+\lip{e + e} or \lip{e < e}.
 These are notionally symmetric, so it would not be appropriate to simply give the left-most one precedence.
@@
 If both synthesize a type with the same type constructor, or only one synthesizes a type, that type constructor is delegated responsibility via a class method, e.g.
-\lstinlinep{syn_BinOp_Add} on line \ref{syn_BinOp_Add} of Listing \ref{string_in}.
+\lip{syn_BinOp_Add} on line \ref{syn_BinOp_Add} of Listing \ref{string_in}.
 
 If both synthesize a type but with different type constructors (e.g.
-we want to concatenate a \lstinlinep{string} and a \lstinlinep{string_in[r]}), then we consult the \emph{handle sets} associated as a class attribute with each type constructor, e.g.
-\lstinlinep{handles_Add_with} on line \ref{handles_Add_with} of Listing \ref{string_in}.
+we want to concatenate a \lip{string} and a \lip{string_in[r]}), then we consult the \emph{handle sets} associated as a class attribute with each type constructor, e.g.
+\lip{handles_Add_with} on line \ref{handles_Add_with} of Listing \ref{string_in}.
 This is a set of other type constructors that the type constructor defining the handle set may potentially support binary operations with.
-When a type constructor is defined, the language checks that if \lstinlinep{tycon2} appears in \lstinlinep{tycon1}'s handler set, then \lstinlinep{tycon1} does \emph{not} appear in \lstinlinep{tycon2}'s handler set.
+When a type constructor is defined, the language checks that if \lip{tycon2} appears in \lip{tycon1}'s handler set, then \lip{tycon1} does \emph{not} appear in \lip{tycon2}'s handler set.
 This is a very simple modular analysis (rather than one that can only be performed at ``link-time''), shown in Listing \ref{check_tycon}, that ensures that the type constructor delegated control over each binary expression is deterministic and  unambiguous without arbitrarily preferring one subexpression position over another.
@@
 Once typechecking is complete, the compiler enters the translation phase.
-%The compiler creates an instance of a class inheriting from \lstinlinep{ace.Target} for use during this phase via the \lstinlinep{init_target} method (line \ref{pybase}.54).
+%The compiler creates an instance of a class inheriting from \lip{ace.Target} for use during this phase via the \lip{init_target} method (line \ref{pybase}.54).
 This object provides methods for code generation and supports features like fresh variable generation, adding imports and so on.
-It's interface is not constrained by @\texttt{lang} (we will see what @\texttt{lang} requires of a target below) and the mechanics of code generation are orthogonal to the  focus of this paper, so we will discuss it relatively abstractly.
+It's interface is not constrained by \texttt{typy} (we will see what \texttt{typy} requires of a target below) and the mechanics of code generation are orthogonal to the  focus of this paper, so we will discuss it relatively abstractly.
 The simplest API would be string-based code generation.
 For the Python target we use here, we generate ASTs.
-The API is based directly on the \lstinlinep{ast} library, with a few additional conveniences just mentioned.
+The API is based directly on the \lip{ast} library, with a few additional conveniences just mentioned.
 This phase follows the same delegation protocol as the typechecking phase.
-Each \lstinlinep{check_}/\lstinlinep{syn_}/\lstinlinep{ana_TermCon} method has a corresponding \lstinlinep{trans_TermCon} method.
+Each \lip{check_}/\lip{syn_}/\lip{ana_TermCon} method has a corresponding \lip{trans_TermCon} method.
 These are all now instance methods, because all indices have been fully determined.
 
-Examples of translation methods for the \lstinlinep{fn} and \lstinlinep{record} type constructors are shown in Listing \ref{trans}.
+Examples of translation methods for the \lip{fn} and \lip{record} type constructors are shown in Listing \ref{trans}.
 The output of translation for our example is discussed in the next subsection.
-Translation methods have access to the context and node, as during typechecking, and must return a translation, which for our purposes, is simply another Python syntax tree (in practice, we offer some additional conveniences beyond \lstinlinep{ast}, such as fresh variable generation and lifting of imports and statements inside expressions to appropriate positions, in the module \lstinlinep{astx} distributed with the language).
+Translation methods have access to the context and node, as during typechecking, and must return a translation, which for our purposes, is simply another Python syntax tree (in practice, we offer some additional conveniences beyond \lip{ast}, such as fresh variable generation and lifting of imports and statements inside expressions to appropriate positions, in the module \lip{astx} distributed with the language).
 Translation methods can assume that the term is well-typed and the typechecking phase saves the type that was delegated control, along with the type that was assigned, as attributes of each term processed by the compiler.
@@
 \begin{lstlisting}
-#class fn(atlang.Type): (cont'd)
+#class fn(typy.Type): (cont'd)
   def trans_FunctionDef(self, ctx, node):
@@
 
-#class record(atlang.Type): (cont'd)
+#class record(typy.Type): (cont'd)
   def trans_Dict(self, ctx, e): 
@@
 
-%Because we are targeting Python directly, most of our \lstinlinep{trans} methods are direct translations (factored out into a helper function).
-The main methods of interest that are not entirely trivial are \lstinlinep{trans_FunctionDef_outer} on lines \ref{pybase}.33-\ref{pybase}-37 and the two translation methods in Listing \ref{record}, which implement the logic described in Sec.
+%Because we are targeting Python directly, most of our \lip{trans} methods are direct translations (factored out into a helper function).
+The main methods of interest that are not entirely trivial are \lip{trans_FunctionDef_outer} on lines \ref{pybase}.33-\ref{pybase}-37 and the two translation methods in Listing \ref{record}, which implement the logic described in Sec.
 \ref{compilation}.
 %
-%Each type constructor must also specify a \lstinlinep{trans_type} method.
+%Each type constructor must also specify a \lip{trans_type} method.
 This is important when targeting a typed language (so that the translations of type annotations can be generated, for example).
 As we will see in the next two sections, this is also critical to ensuring type safety.
-The language \emph{checks} not only that translations having a given type are well-typed, but that they have the type specified by this method (requiring that the target provide a \lstinlinep{is_of_type} method).
-Here, because we are simply targeting Python directly, the \lstinlinep{trans_type} method on line \ref{record}.49-\ref{record}.50 simply generates \lstinlinep{target.dyn}.
-%
-%When this phase is complete, each node processed by the context will have a translation, available via the \lstinlinep{trans} attribute.
+The language \emph{checks} not only that translations having a given type are well-typed, but that they have the type specified by this method (requiring that the target provide a \lip{is_of_type} method).
+Here, because we are simply targeting Python directly, the \lip{trans_type} method on line \ref{record}.49-\ref{record}.50 simply generates \lip{target.dyn}.
+%
+%When this phase is complete, each node processed by the context will have a translation, available via the \lip{trans} attribute.
 In particular, each typed function has a translation.
@@
 %
-%To support external compilation, the target must have an \lstinlinep{emit} method that takes the compilation script's file name and a reference to a string generator (an instance of \lstinlinep{ace.util.CG}) and emits source code.
+%To support external compilation, the target must have an \lip{emit} method that takes the compilation script's file name and a reference to a string generator (an instance of \lip{ace.util.CG}) and emits source code.
 The string generator we provide can track indentation levels (to support Python code generation and make generation for other languages more readable, for the purposes of debugging).
@@
 \begin{lstlisting}
-import atlib.runtime as _at_i0_
+import typy.std.runtime as _at_i0_
 
@@
 
-Listing \ref{external-compilation} shows how to invoke the \lstinlinep{@} compiler at the shell to typecheck and translate then execute \lstinlinep{listing1.py}.
+Listing \ref{external-compilation} shows how to invoke the \lip{@} compiler at the shell to typecheck and translate then execute \lip{listing1.py}.
 %Note that the {\texttt{print}} statements at the top  of the  compilation script were evaluated during compilation only.
-%These two steps can be combined by running \lstinlinep{ace listing1.py} (the intermediate file is not generated unless explicitly requested in this case).
-The @\texttt{lang} compiler is itself a Python library and \lstinlinep{@} is a simple Python script that invokes it in two steps:
+%These two steps can be combined by running \lip{ace listing1.py} (the intermediate file is not generated unless explicitly requested in this case).
+The \texttt{typy} compiler is itself a Python library and \lip{@} is a simple Python script that invokes it in two steps:
 \begin{enumerate}
 \item It evaluates the compilation script to completion.
-\item For any top-level bindings in the environment that are @\texttt{lang} functions, it initiates typehecking and translation as described above.
+\item For any top-level bindings in the environment that are \texttt{typy} functions, it initiates typehecking and translation as described above.
 If no type errors are discovered, the ordered set of translations are collected (obeying order dependencies) and emitted.
@@
 \begin{enumerate}
-\item \lstinlinep{fn} recognized the function name \lstinlinep{__main__} as special, inserting the standard Python code to invoke it if the file is run directly.
+\item \lip{fn} recognized the function name \lip{__main__} as special, inserting the standard Python code to invoke it if the file is run directly.
 \item Records translated to tuples (the field names were not needed).
 \item Decimals translated to pairs of integers. String conversion is defined in a ``runtime'' package with a ``fresh'' name.
-\item Terms of type \lstinlinep{string_in[r"..."]} translated to strings.
+\item Terms of type \lip{string_in[r"..."]} translated to strings.
 Checks were all performed  statically in this example.% Note that there are situations (e.g.
@@
 If analagous code were written in Python itself, these could only be found if the code was executed on the first day of the month (and not all of the issues would  immediately be caught as run-time exceptions).
-In @\texttt{lang}, these can be found without executing the code (i.e.
+In \texttt{typy}, these can be found without executing the code (i.e.
 we have true static typechecking).
@@
   File "listing^\ref{oops}^", line 9, in <module>
-atlang.TypeError: 
+typy.TypeError: 
     File "listing^\ref{oops}^.py", line ^\ref{error-start}^, col 12, in <module>: 
@@
 \end{lstlisting}
-\caption{Execution never proceeds into a function with a type error when using @$\texttt{lang}$ for implicit compilation.}
+\caption{Execution never proceeds into a function with a type error when using \texttt{typy} for implicit compilation.}
 \label{oops-exec}
@@
 
-\subsection{Interactive Invocation}\label{interactive} @\texttt{lang} functions over values of type \lstinlinep{dyn} can be invoked directly from Python.
+\subsection{Interactive Invocation}\label{interactive} \texttt{typy} functions over values of type \lip{dyn} can be invoked directly from Python.
 Typechecking and translation occurs upon first invocation (subsequent calls are cached; we assume that the static environment is immutable).
@@
 If a function captures any static values that do not support lifting, then that function can only be used via interactive invocation (there is no way to separate compile-time from run-time without lifting captured values).
-Such values when captured in an interactively invoked setting by default synthesize type \lstinlinep{dyn}, though they can provide a \lstinline{atlang_syn_Capture} method that synthesizes a more specific type.
-We will see an example of this and a related use of phaseless capture in our discussion of \lstinlinep{atlib.opencl} below.
-
-%\todo{move to future work section?}In future work, we plan to detail how to insert dynamic checks and wrapper objects, defined by active type constructors in a manner similar to how static checks are defined here, so that values that are not of type \lstinlinep{dyn} can be passed in and out of untyped Python code.
-Because these can be seen as implicit coercions to and from \lstinlinep{dyn}, and we do not aim to introduce this feature in this paper, we omit discussion of this topic.
+Such values when captured in an interactively invoked setting by default synthesize type \lip{dyn}, though they can provide a \lip{typy_syn_Capture} method that synthesizes a more specific type.
+We will see an example of this and a related use of phaseless capture in our discussion of \lip{typy.std.opencl} below.
+
+%\todo{move to future work section?}In future work, we plan to detail how to insert dynamic checks and wrapper objects, defined by active type constructors in a manner similar to how static checks are defined here, so that values that are not of type \lip{dyn} can be passed in and out of untyped Python code.
+Because these can be seen as implicit coercions to and from \lip{dyn}, and we do not aim to introduce this feature in this paper, we omit discussion of this topic.
 Explicit coercions can be implemented using the mechanisms described above.
-For example, \lstinlinep{string_in} provides an introductory form that checks a provided \lstinlinep{string}  or \lstinlinep{dyn} value dynamically, raising an exception in the case of failure: \lstinlinep{[raw_input()] : string_in[r"\\d+"]}.
-
-%.; for example, the unexpected use of \lstinlinep{None}\footnote{The notation \texttt{+T} constructs a \texttt{None}-able option type; see appendix.}).
+For example, \lip{string_in} provides an introductory form that checks a provided \lip{string}  or \lip{dyn} value dynamically, raising an exception in the case of failure: \lip{[raw_input()] : string_in[r"\\d+"]}.
+
+%.; for example, the unexpected use of \lip{None}\footnote{The notation \texttt{+T} constructs a \texttt{None}-able option type; see appendix.}).
 Static type checking allows us to find these errors during compilation.
@@
 The compilation script completes (so that functions can refer to each other mutually recursively), then the typed functions in the top-level environment are typechecked.
-The typechecker raises an exception statically at the first error, and \lstinlinep{acec} prints it to the console as shown.
+The typechecker raises an exception statically at the first error, and \lip{acec} prints it to the console as shown.
 
 %When encountering a compound term (e.g.
-\lstinlinep{t.amount} in Listing \ref{example}), the compiler defers control over type  and translation to the type of a designated subexpression (e.g.
-\lstinlinep{t}) according to a fixed fixed \emph{dispatch protocol}.
-Below are examples of the choices made in @\texttt{lang}.\todo{full in appendix?}
+\lip{t.amount} in Listing \ref{example}), the compiler defers control over type  and translation to the type of a designated subexpression (e.g.
+\lip{t}) according to a fixed fixed \emph{dispatch protocol}.
+Below are examples of the choices made in \texttt{typy}.\todo{full in appendix?}
 %\begin{itemize}
-%\item Responsibility over {\bf attribute access} (\texttt{e.attr}), {\bf subscripting} (\texttt{e[e1]}) and \textbf{calls} (\lstinlinep{e(e1, ..., en)}) and {\bf unary operations} (e.g.
-\lstinlinep{-e}) is handed to the type recursively assigned to \texttt{e}.
+%\item Responsibility over {\bf attribute access} (\texttt{e.attr}), {\bf subscripting} (\texttt{e[e1]}) and \textbf{calls} (\lip{e(e1, ..., en)}) and {\bf unary operations} (e.g.
+\lip{-e}) is handed to the type recursively assigned to \texttt{e}.
 %\item Responsibility over {\bf binary operations} (e.g.
-\lstinlinep{e1 + e2}) is first handed to the type assigned to the left operand.
+\lip{e1 + e2}) is first handed to the type assigned to the left operand.
 If it indicates that it cannot handle the operation, the type assigned to the right operand is handed responsibility.
@@
 \ref{related} for a discussion.}
-%\item Responsibility over \textbf{constructor calls} (\lstinlinep{[t](e1, ..., en)}), where \lstinlinep{t} is a \emph{compile-time Python expression} evaluating to an active type, is handed to that type.
-If \lstinlinep{t} evaluates to a \emph{family} of types, like \lstinlinep{clx.Cplx}, the active type is first generated via a class method, as discussed below.
+%\item Responsibility over \textbf{constructor calls} (\lip{[t](e1, ..., en)}), where \lip{t} is a \emph{compile-time Python expression} evaluating to an active type, is handed to that type.
+If \lip{t} evaluates to a \emph{family} of types, like \lip{clx.Cplx}, the active type is first generated via a class method, as discussed below.
 %%\item Responsibility over {\bf simple assignment statements}, is handed to the type of the variable on the left (which, as we will see below, is determined by the active base).
@@
 %%
-%%\todo{To be revised from here on down}The core of @\texttt{lang} consists of about 1500 lines of Python code implementing its primary concepts: generic functions, concrete functions, active types, active bases and active targets.
-The latter three comprise @\texttt{lang}'s extension mechanism.
-Extensions provide semantics to, and govern the compilation of, @\texttt{lang} functions, rather than logic in @\texttt{lang}'s core.
-%Indeed, the name ``@\texttt{lang}'' might itself be an acronym: it is an ``active compilation environment''.
-
-% As a result, execution is essentially as fast as it could be given the target language chosen (Python). We will target an even faster language, \lstinlinep{OpenCL}, in Sec. \ref{targets}.
-
-%Because \lstinlinep{t} will be statically guaranteed to have type \lstinlinep{Transfer},
+%%\todo{To be revised from here on down}The core of \texttt{typy} consists of about 1500 lines of Python code implementing its primary concepts: generic functions, concrete functions, active types, active bases and active targets.
+The latter three comprise \texttt{typy}'s extension mechanism.
+Extensions provide semantics to, and govern the compilation of, \texttt{typy} functions, rather than logic in \texttt{typy}'s core.
+%Indeed, the name ``\texttt{typy}'' might itself be an acronym: it is an ``active compilation environment''.
+
+% As a result, execution is essentially as fast as it could be given the target language chosen (Python). We will target an even faster language, \lip{OpenCL}, in Sec. \ref{targets}.
+
+%Because \lip{t} will be statically guaranteed to have type \lip{Transfer},
 
 \section{More Examples}\label{examples}
-In the previous section, we showed examples of several interesting type system fragments implemented as libraries using \lstinlinep{atlang}, including functional record types, a prototypic object system and regular string types.
-A more detailed description of regular string types and their implementation using \lstinlinep{atlang} was recently published at a workshop \cite{sanitation-psp14}.
+In the previous section, we showed examples of several interesting type system fragments implemented as libraries using \lip{typy}, including functional record types, a prototypic object system and regular string types.
+A more detailed description of regular string types and their implementation using \lip{typy} was recently published at a workshop \cite{sanitation-psp14}.
 Here, we showcase two more powerful examples that demonstrate the expressive power of the system: functional datatypes with nested pattern matching, and a type safe foreign function interface to the complete OpenCL language for many-core programming on devices like GPUs.
@@
 \begin{lstlisting}
-from atlang import ty
-from atlib import case, casetype, tuple, fn
+from typy import ty
+from typy.std import case, casetype, tuple, fn
 
@@
 
-class caseop(atlang.Type):
+class caseop(typy.Type):
   def syn_Call(self, ctx, e):
@@
     if len(e.args) != 0:
-      raise atlang.TypeError("...", e)
+      raise typy.TypeError("...", e)
     return caseop_applied[()]
 
-class caseop_applied(atlang.Type):
+class caseop_applied(typy.Type):
   def syn_Subscript(self, ctx, e):
@@
       if ty is not None and cur_type != ty:
-        raise atlang.TypeError("inconsistent", rule)
+        raise typy.TypeError("inconsistent", rule)
       else: ty = cur_type
@@
 
-class casetype(atlang.Type):
+class casetype(typy.Type):
   # ...
@@
 
-class tuple(atlang.Type):
+class tuple(typy.Type):
   # ...
@@
 For example, tree structures are well-modeled using recursive labeled sums, as we show in Listing \ref{example-casetypes}.
-Here, \lstinlinep{Tree} is a \emph{case type}, which is what we call labeled sums in \lstinlinep{atlib}.
-\lstinlinep{Tree} is declared as a \emph{named type} using the \lstinlinep{@ty} annotation.
-This is a general mechanism in \lstinlinep{atlang} for supporting recursive types that behave generatively, i.e.
+Here, \lip{Tree} is a \emph{case type}, which is what we call labeled sums in \lip{typy.std}.
+\lip{Tree} is declared as a \emph{named type} using the \lip{@ty} annotation.
+This is a general mechanism in \lip{typy} for supporting recursive types that behave generatively, i.e.
 that are identified by a name.
 Functional datatypes in ML and similar languages are also generative in this way.
-The \lstinlinep{@ty} mechanism also supports parameterized types: here, \lstinlinep{a} stands in for any other type.
-The type index to \lstinlinep{casetype} is a series of cases with, optionally, associated types.
-Here, a (binary) tree can either be empty (no associated data), a leaf with an associated value of type \lstinlinep{a}, or an internal node, which takes a pair of trees as data.
-We use the \lstinlinep{atlib.tuple} type constructor to represent pairs.
-
-The function \lstinlinep{treesum} on lines 11-17 takes a tree containing \lstinlinep{dyn} values and computes the sum of these values by nested pattern matching.
-The \lstinlinep{__main__} function constructs a tree and calls \lstinlinep{treesum} on it.
+The \lip{@ty} mechanism also supports parameterized types: here, \lip{a} stands in for any other type.
+The type index to \lip{casetype} is a series of cases with, optionally, associated types.
+Here, a (binary) tree can either be empty (no associated data), a leaf with an associated value of type \lip{a}, or an internal node, which takes a pair of trees as data.
+We use the \lip{typy.std.tuple} type constructor to represent pairs.
+
+The function \lip{treesum} on lines 11-17 takes a tree containing \lip{dyn} values and computes the sum of these values by nested pattern matching.
+The \lip{__main__} function constructs a tree and calls \lip{treesum} on it.
 The result of compiling Listing \ref{example-casetypes} is shown in Listing \ref{example-casetypes-out}.
@@
 Some relevant portions of the implementation of case analysis are shown in Listing \ref{example-casetypes-impl}.
-The object \lstinlinep{case} is an instance of the class \lstinlinep{_case}, which defines the method \lstinlinep{syn_Lift}.
-When the imported name \lstinlinep{case} is encountered in Listing \ref{example-casetypes}, this is called, per the discussion earlier on names in the static environment.
-The result is that \lstinline{case} synthesizes type \lstinlinep{caseop[()]}.
-
-The only purpose of the type constructor \lstinlinep{caseop} is to define the \lstinlinep{syn_Call} method, which ensures that the case operator is applied to a single well-typed scrutinee.
-The combined term \lstinlinep{case(tree)} then synthesizes type \lstinlinep{caseop_applied[()]}.
-Notice that neither method is paired with a \lstinlinep{trans_} method -- these types are merely ``bookkeeping'' for the compound case analysis form.
-Parts of this form are not meaningful by themselves, so translation would fail if, for example, \lstinlinep{case(tree)} was written by itself without cases.
-
-The \lstinlinep{caseop_applied} tycon defines the \lstinlinep{syn_Subscript} method, which actually processes the list of provided rules.
+The object \lip{case} is an instance of the class \lip{_case}, which defines the method \lip{syn_Lift}.
+When the imported name \lip{case} is encountered in Listing \ref{example-casetypes}, this is called, per the discussion earlier on names in the static environment.
+The result is that \lip{case} synthesizes type \lip{caseop[()]}.
+
+The only purpose of the type constructor \lip{caseop} is to define the \lip{syn_Call} method, which ensures that the case operator is applied to a single well-typed scrutinee.
+The combined term \lip{case(tree)} then synthesizes type \lip{caseop_applied[()]}.
+Notice that neither method is paired with a \lip{trans_} method -- these types are merely ``bookkeeping'' for the compound case analysis form.
+Parts of this form are not meaningful by themselves, so translation would fail if, for example, \lip{case(tree)} was written by itself without cases.
+
+The \lip{caseop_applied} tycon defines the \lip{syn_Subscript} method, which actually processes the list of provided rules.
 First, the type of the scrutinee is asked to process the bindings from the pattern portion of each rule.
-This involves calling methods of the form \lstinlinep{type_pat_TermCon}, which must return types for all the sub-patterns.
-For example, the \lstinline{type_pat_Call} method of \lstinline{casetype} makes sure that the case named is defined, and if so returns the associated type, taken from its type index, as the type of the inner pattern.
-The process repeats recursively, so that a tuple pattern might then be processed by \lstinlinep{type_pat_Tuple}.
+This involves calling methods of the form \lip{type_pat_TermCon}, which must return types for all the sub-patterns.
+For example, the \lip{type_pat_Call} method of \lip{casetype} makes sure that the case named is defined, and if so returns the associated type, taken from its type index, as the type of the inner pattern.
+The process repeats recursively, so that a tuple pattern might then be processed by \lip{type_pat_Tuple}.
 Any type constructor can participate in this protocol by defining appropriate methods, so it is extensible.
-Note, however, that this is a ``second-order'' extensibility mechanism -- nothing in \verb|atlang| itself supports pattern matching.
-Instead, \lstinlinep{case} itself defined this protocol in a library, demonstrating the power of using a flexible general-purpose language for writing extensions.
-
-The translation phase begins at the \lstinlinep{trans_Subscript} method of \lstinlinep{caseop_applied}.
+Note, however, that this is a ``second-order'' extensibility mechanism -- nothing in \verb|typy| itself supports pattern matching.
+Instead, \lip{case} itself defined this protocol in a library, demonstrating the power of using a flexible general-purpose language for writing extensions.
+
+The translation phase begins at the \lip{trans_Subscript} method of \lip{caseop_applied}.
 This simply hands responsibility for translating the entire set of rules to the type of the scrutinee.
@@
 
-By combining the functional behavior of the \lstinlinep{fn} type constructor, which dispenses with ``return'' statements, and creatively repurposing existing syntax, we are thus able to implement an essentially idiomatic statically-typed functional language, entirely as a library for Python.
+By combining the functional behavior of the \lip{fn} type constructor, which dispenses with ``return'' statements, and creatively repurposing existing syntax, we are thus able to implement an essentially idiomatic statically-typed functional language, entirely as a library for Python.
 The constructs we defined can be composed arbitrarily with, e.g.
@@
 \begin{lstlisting}
-import atlib.opencl as opencl
+import typy.std.opencl as opencl
 import numpy
@@
 \end{lstlisting}
-\caption{The underlying code generated by \texttt{atlib.opencl} as a string passed through \texttt{pyopencl}.}
+\caption{The underlying code generated by \texttt{typy.std.opencl} as a string passed through \texttt{pyopencl}.}
 \label{example-opencl-out}
@@
 \begin{lstlisting}
-class template_fn(atlang.Type):
+class template_fn(typy.Type):
   # ...
@@
   # ...
-  def atlang_syn_Capture(self):
+  def typy_syn_Capture(self):
     return global_ptr[
@@
 GPUs) via a standard API, as well as a runtime that allows a program to manage device memory.
-The \lstinlinep{pyopencl} package exposes these APIs to Python code and integrates them with the popular \lstinlinep{numpy} numeric array package \cite{klockner2011pycuda}.
+The \lip{pyopencl} package exposes these APIs to Python code and integrates them with the popular \lip{numpy} numeric array package \cite{klockner2011pycuda}.
 To compile an OpenCL function, however, users must write it as a string.
@@
 
-In Listing \ref{example-opencl}, we show usage of the \lstinlinep{atlib.opencl} package, which implements the entirety of the OpenCL type system (which includes essentially the entirety of C99, plus some extensions to work with multiple memory spaces).
+In Listing \ref{example-opencl}, we show usage of the \lip{typy.std.opencl} package, which implements the entirety of the OpenCL type system (which includes essentially the entirety of C99, plus some extensions to work with multiple memory spaces).
 Although the details are beyond the scope of this paper, we note that the mapping onto Python syntax was straightforward, particularly given the flexibility of analytic numeric literal forms, as described above, to support the wide variety of number types in C99-based languages.
 
-After initializing a device, line 4 sends a \lstinlinep{numpy} array to the device, assigning \lstinlinep{buffer} its handle, an object inheriting from \lstinlinep{opencl.Buffer}.
-The type of the \lstinlinep{numpy} array is also extracted and stored in this object.
+After initializing a device, line 4 sends a \lip{numpy} array to the device, assigning \lip{buffer} its handle, an object inheriting from \lip{opencl.Buffer}.
+The type of the \lip{numpy} array is also extracted and stored in this object.
 
 On lines 7-18, we define three OpenCL template function.
-Let us consider the first, \lstinlinep{map}, which takes two arguments.
+Let us consider the first, \lip{map}, which takes two arguments.
 A template function is one that does not have specified argument types, generally because many types might be valid, as in this case.
-The body of \lstinlinep{map} would be valid for any combination of types for \lstinlinep{b} and \lstinlinep{f} that support subscripting and calls as shown.
+The body of \lip{map} would be valid for any combination of types for \lip{b} and \lip{f} that support subscripting and calls as shown.
 Template functions thus defer typechecking to every invocation site, based on the types synthesized by the provided arguments.
@@
 
-We see two applications of \lstinlinep{map} on lines 20 and 21.
+We see two applications of \lip{map} on lines 20 and 21.
 Note that these are in the dynamically typed portion of the code.
-By default, the first argument in each case, \lstinlinep{buffer}, would synthesize type \lstinlinep{dyn} and typechecking would fail (at line 10, technically, because \lstinlinep{dyn} doesn't support subscripting with the type that \lstinlinep{gid} synthesizes).
+By default, the first argument in each case, \lip{buffer}, would synthesize type \lip{dyn} and typechecking would fail (at line 10, technically, because \lip{dyn} doesn't support subscripting with the type that \lip{gid} synthesizes).
 But we make use of the support for phaseless capture discussed in Sec.
-\ref{interactive}, which allows values to define a method, \lstinlinep{atlang_syn_Capture}, that permits capture and, in this case, function application with values that would not otherwise support crossing into \lstinlinep{atlang} functions with argument types other than \lstinlinep{dyn}.
-In this case, \lstinlinep{Buffer} objects synthesize a corresponding pointer type, represented by the tycon \lstinlinep{opencl.global_ptr} and indexed by the target type of the pointer, here \lstinlinep{opencl.double} because the initial \lstinlinep{numpy} array was an array of \lstinlinep{double}s and there is a mapping from \lstinlinep{numpy} data types to OpenCL types.
+\ref{interactive}, which allows values to define a method, \lip{typy_syn_Capture}, that permits capture and, in this case, function application with values that would not otherwise support crossing into \lip{typy} functions with argument types other than \lip{dyn}.
+In this case, \lip{Buffer} objects synthesize a corresponding pointer type, represented by the tycon \lip{opencl.global_ptr} and indexed by the target type of the pointer, here \lip{opencl.double} because the initial \lip{numpy} array was an array of \lip{double}s and there is a mapping from \lip{numpy} data types to OpenCL types.
 The implementation is shown in Listing \ref{example-opencl-buffer}.
@@
 
-Each invocation of \lstinlinep{map} received the same buffer, but a different function.
+Each invocation of \lip{map} received the same buffer, but a different function.
 Because these were template functions, their types were distinct.
 As such, two different versions of map were generated, as shown in Listing \ref{example-opencl-out}.
-Indeed, without support for function pointers in OpenCL, code generation like this is the standard way of supporting ``higher-order'' functions like \lstinlinep{map}.
+Indeed, without support for function pointers in OpenCL, code generation like this is the standard way of supporting ``higher-order'' functions like \lip{map}.
 Rather than having to do this manually, however, our type system handled this internally.
@@
 Static terms have a \emph{static dynamics} and evaluate to \emph{static values} or fail.
-Static terms are analagous to top-level Python code in \verb|atlang|, and external terms are analagous to terms inside typed functions.
+Static terms are analagous to top-level Python code in \verb|typy|, and external terms are analagous to terms inside typed functions.
 
@@
 \ref{nat-sugared}.
-In @\texttt{lang},  fragment definitions are packaged separately; we do not include this in the core calculus for simplicity.
+In \texttt{typy},  fragment definitions are packaged separately; we do not include this in the core calculus for simplicity.
 %
@@
 
-%In this section, we will develop an ``actively typed'' version of the simply-typed lambda calculus with simply-kinded type-level computation called $\lam@\texttt{lang}$.
+%In this section, we will develop an ``actively typed'' version of the simply-typed lambda calculus with simply-kinded type-level computation called $\lam\texttt{typy}$.
 More specifically, the level of types, $\tau$, will itself form a simply-typed lambda calculus.
 \emph{Kinds} classify type-level terms in the same way that types conventionally classify expressions.
-Types become just one kind  of type-level value (which we will write $\kTypeBlur$, though it is also variously written $\star$, \lstinlinep{T} and \lstinlinep{Type} in various settings).
+Types become just one kind  of type-level value (which we will write $\kTypeBlur$, though it is also variously written $\star$, \lip{T} and \lip{Type} in various settings).
 Rather than there being a fixed set of type and operator constructors, we allow the programmer to declare new  constructors, and give their static and dynamic semantics by writing type-level functions.
@@
 %
-%The syntax of Core $\lam@\texttt{lang}$ is given in Fig.
+%The syntax of Core $\lam\texttt{typy}$ is given in Fig.
 \ref{grammar}.
-An example of a program defining type and operator constructors that can be used to construct an active embedding of G\"odel's \textbf{T} into $\lam@\texttt{lang}$ is given in Fig.
+An example of a program defining type and operator constructors that can be used to construct an active embedding of G\"odel's \textbf{T} into $\lam\texttt{typy}$ is given in Fig.
 \ref{nat}.
@@
 %\end{flalign}
-%\caption{The definition of the record type in $\lam@\texttt{lang}$ using the desugarings in Figure \ref{desugaring} and with the addition of simple let bindings for both type and term variables (not shown).
+%\caption{The definition of the record type in $\lam\texttt{typy}$ using the desugarings in Figure \ref{desugaring} and with the addition of simple let bindings for both type and term variables (not shown).
 Some type-level helper functions also omitted for concision.}
@@
 %\]
-%This anchors our exposition; we will describe how it is derived (i.e. how to write a compiler for $\lam@\texttt{lang}$) in the following sections. 
+%This anchors our exposition; we will describe how it is derived (i.e. how to write a compiler for $\lam\texttt{typy}$) in the following sections. 
 The static semantics are specified by the \emph{bidirectional active typechecking and translation judgements} shown in Fig.
@@
 Failure cases for the static helper functions evaluate to $\FF{fail}$.
-Derivation of the typing judgement does not continue if a $\FF{fail}$ occurs (corresponding to an \lstinlinep{atlang.TypeError} propagated directly to the compiler).
+Derivation of the typing judgement does not continue if a $\FF{fail}$ occurs (corresponding to an \lip{typy.TypeError} propagated directly to the compiler).
 
@@
 \ref{tleval}).
-This function can be compared to the method \lstinlinep{ana_Dict} in Listing \ref{record}.
+This function can be compared to the method \lip{ana_Dict} in Listing \ref{record}.
 
@@
 The type index and translation are recursively formed.
-This can be compared to the class method \lstinlinep{syn_idx_Dict} in Listing \ref{record}.
+This can be compared to the class method \lip{syn_idx_Dict} in Listing \ref{record}.
 
@@
 $\FF{let}~\{\texttt{a}{=}x, \texttt{b}{=}y\}=r~\FF{in}~e$.
-@$\texttt{lang}$ supports this using Python's syntax for destructuring assignment, but we must omit the details.
+\verb|typy| supports this using Python's syntax for destructuring assignment, but we must omit the details.
 
@@
 The classifiers of type-level terms are called \emph{kinds}, $\kappa$, to distinguish them from  \emph{types}, which are  type-level values of kind $\kTypeBlur$.
-%In @\texttt{lang}, the type-level language (together with the level of programs) is written in Python, which can be thought of as having a rather limited kind system (with one kind, \lstinlinep{dyn}).
+%In \texttt{typy}, the type-level language (together with the level of programs) is written in Python, which can be thought of as having a rather limited kind system (with one kind, \lip{dyn}).
 Here, we are able to more precisely discuss the kinds of values in the type-level language.
@@
 %
-%%We will write the kind of types as $\kTypeBlur$, though it is also written $\star$ or \lstinlinep{Type} in various similarly structured languages (see Sec. \ref{related-work}). 
+%%We will write the kind of types as $\kTypeBlur$, though it is also written $\star$ or \lip{Type} in various similarly structured languages (see Sec. \ref{related-work}). 
 %% Rather than there being a fixed set of type constructors, we allow the programmer to declare new type  constructors, and give the static and dynamic semantics of their associated operators, by writing type-level functions.
@@
 %The syntax for external terms (Figure \ref{grammar}) contains variables, $\lambda$ terms, three generalized introductory forms and a single generalized elimination form.
-The introductory forms are either unascribed, ascribed with a type or ascribed with a type constructor, as in our discussion of @\texttt{lang}.
+The introductory forms are either unascribed, ascribed with a type or ascribed with a type constructor, as in our discussion of \texttt{typy}.
 These generalized forms take a single a type-level value as an {index} and $n \geq 0$ arguments, which are other external terms.
 To better motivate this choice, we can give a purely syntactic desugaring of a Python-like syntax with labels to these forms, shown in Figure \ref{desugaring}.
-It is instructive to rewrite lines 27-28 of Figure \ref{record-theory} using these desugarings.% In @\texttt{lang}, desugarings can be user-defined
+It is instructive to rewrite lines 27-28 of Figure \ref{record-theory} using these desugarings.% In \texttt{typy}, desugarings can be user-defined
 %
@@
 In our first example, all the operator constructors are indexed trivially (by index kind $\kunit$), but other examples use more interesting indices.
-For example, in SML, the projection operator \lstinlinep{#3} can be applied to an $n$-tuple, $e$, iff $n \geq 3$.
+For example, in SML, the projection operator \lip{#3} can be applied to an $n$-tuple, $e$, iff $n \geq 3$.
 Note that it thus cannot be a function with a standard arrow type.
-Notionally, \lstinlinep{#} is an operator constructor and \lstinlinep{3} is its index.
-In an active embedding of $n$-tuples into $\lam@\texttt{lang}$, this would be written $\eop{Tuple}{prj}{3}{e}$ (we will nearly recover ML's syntax later).
+Notionally, \lip{#} is an operator constructor and \lip{3} is its index.
+In an active embedding of $n$-tuples into $\lam\texttt{typy}$, this would be written $\eop{Tuple}{prj}{3}{e}$ (we will nearly recover ML's syntax later).
 $\fvar{LabeledTuple}.\opvar{prj}$ is the operator constructor used to access a field of a labeled tuple, so it has index kind $\klabel$.
@@
 In the core calculus, this is one of three functions, called \FF{iana}, \FF{isyn} and \FF{esyn} in the grammar.\footnote{This means that any particular type constructor only supports a single introduction and elimination desugaring.
-This is a minor inconvenience in some cases that is resolved in @\texttt{lang} by the use of methods.} We see how this is done with the rules for the active typing judgements, given in Figure \ref{atj}.
+This is a minor inconvenience in some cases that is resolved in \texttt{typy} by the use of methods.} We see how this is done with the rules for the active typing judgements, given in Figure \ref{atj}.
 
@@
 Given such a type, the rule extracts a definition, named $\FF{iana}$, from the type constructor (cf.
-the \lstinlinep{ana_Dict} and \lstinlinep{trans_Dict} methods earlier).
+the \lip{ana_Dict} and \lip{trans_Dict} methods earlier).
 It calls this function with a pair containing the operator and type index and a list of \emph{reified arguments}, which have kind $\Q$ and introductory form $(\Gamma; e)?$.
@@
 %%\vspace{-10pt}
-%\caption{\small Syntax of Core $\lam@\texttt{lang}$.
+%\caption{\small Syntax of Core $\lam\texttt{typy}$.
 Here, $x$ ranges over external and internal language variables, $\tvar{t}$ ranges over type-level variables, $\fvar{tycon}$ ranges over type constructor names and $\ell$ ranges over labels.
@@
 %\end{figure}
-%We will now give a core typed lambda calculus, $\lam@\texttt{lang}$, that captures the semantics described in the previous sections.
+%We will now give a core typed lambda calculus, $\lam\texttt{typy}$, that captures the semantics described in the previous sections.
 It is intended to make precise how active typechecking and translation works and how our mechanism relates to existing work on bidirectional typechecking, type-level computation and typed compilation while abstracting away from the details of Python's syntax and imposing a stronger type-level semantics that will allow us to state metatheoretic properties of interest.
@@
 %
-%The syntax of $\lam@\texttt{lang}$ is shown in Figure \ref{grammar} and an example \emph{program} that defines a type constructor, $\fvar{record}$, with a semantics similar to that given in Listing \ref{record}\footnote{Technically, this defines labeled tuples, because the order of labels matters.}, using it to write an identity function and compute the value $triv$ (the empty record), is shown in Figure \ref{record-theory}.
-%
-%%In this section, we will develop an ``actively typed'' version of the simply-typed lambda calculus with simply-kinded type-level computation called $\lam@\texttt{lang}$.
+%The syntax of $\lam\texttt{typy}$ is shown in Figure \ref{grammar} and an example \emph{program} that defines a type constructor, $\fvar{record}$, with a semantics similar to that given in Listing \ref{record}\footnote{Technically, this defines labeled tuples, because the order of labels matters.}, using it to write an identity function and compute the value $triv$ (the empty record), is shown in Figure \ref{record-theory}.
+%
+%%In this section, we will develop an ``actively typed'' version of the simply-typed lambda calculus with simply-kinded type-level computation called $\lam\texttt{typy}$.
 More specifically, the level of types, $\tau$, will itself form a simply-typed lambda calculus.
 \emph{Kinds} classify type-level terms in the same way that types conventionally classify expressions.
-Types become just one kind  of type-level value (which we will write $\kTypeBlur$, though it is also variously written $\star$, \lstinlinep{T} and \lstinlinep{Type} in various settings).
+Types become just one kind  of type-level value (which we will write $\kTypeBlur$, though it is also variously written $\star$, \lip{T} and \lip{Type} in various settings).
 Rather than there being a fixed set of type and operator constructors, we allow the programmer to declare new  constructors, and give their static and dynamic semantics by writing type-level functions.
@@
 %%
-%%The syntax of Core $\lam@\texttt{lang}$ is given in Fig.
+%%The syntax of Core $\lam\texttt{typy}$ is given in Fig.
 \ref{grammar}.
-An example of a program defining type and operator constructors that can be used to construct an active embedding of G\"odel's \textbf{T} into $\lam@\texttt{lang}$ is given in Fig.
+An example of a program defining type and operator constructors that can be used to construct an active embedding of G\"odel's \textbf{T} into $\lam\texttt{typy}$ is given in Fig.
 \ref{nat}.
@@
 %%\]
-%%This anchors our exposition; we will describe how it is derived (i.e. how to write a compiler for $\lam@\texttt{lang}$) in the following sections. 
+%%This anchors our exposition; we will describe how it is derived (i.e. how to write a compiler for $\lam\texttt{typy}$) in the following sections. 
 %The key judgements are the \emph{bidirectional active typing judgements}  (Fig.
@@
 %
-%In $\lam@\texttt{lang}$, the internal language (IL) provides partial functions (via the generic fixpoint operator of Plotkin's PCF) and simple product  types for the sake of our example.
+%In $\lam\texttt{typy}$, the internal language (IL) provides partial functions (via the generic fixpoint operator of Plotkin's PCF) and simple product  types for the sake of our example.
 In practice, the internal language could be any typed  language with a specification for which type safety and decidability of typechecking have been satisfyingly determined.
@@
 %\subsection{Types and Type-Level Computation}\label{types}
-%$\lam@\texttt{lang}$ supports, and makes extensive use of, simply-kinded type-level computation.
+%$\lam\texttt{typy}$ supports, and makes extensive use of, simply-kinded type-level computation.
 Specifically, type-level terms, $\tau$, themselves form a typed lambda calculus.
@@
 Types are  type-level values of kind $\kTypeBlur$.
-In @\texttt{lang}, the type-level language (together with the level of programs) is written in Python, which can be thought of as having a rather limited kind system (with one kind, \lstinlinep{dyn}).
+In \texttt{typy}, the type-level language (together with the level of programs) is written in Python, which can be thought of as having a rather limited kind system (with one kind, \lip{dyn}).
 Here, we are able to more precisely discuss the kinds of values in the type-level language.
@@
 %In most languages, types are formed by applying one of a collection of \emph{type constructors} to zero or more \emph{indices}.
-In $\lam@\texttt{lang}$, the situation is notionally similar.
+In $\lam\texttt{typy}$, the situation is notionally similar.
 User-defined type constructors can be declared at the top of a program (or lifted to the top, in practice) using \textsf{tycon}.
@@
 %
-%Every type constructor also defines a \emph{representation}, a type-level function that associates with every type an internal \emph{representation type} (analagous to \lstinlinep{trans_type} above).
+%Every type constructor also defines a \emph{representation}, a type-level function that associates with every type an internal \emph{representation type} (analagous to \lip{trans_type} above).
 We will return to this after introducing the external forms and operator definitions.
@@
 %The syntax for external terms (Figure \ref{grammar}) contains variables, $\lambda$ terms, three generalized introductory forms and a single generalized elimination form.
-The introductory forms are either unascribed, ascribed with a type or ascribed with a type constructor, as in our discussion of @\texttt{lang}.
+The introductory forms are either unascribed, ascribed with a type or ascribed with a type constructor, as in our discussion of \texttt{typy}.
 These generalized forms take a single a type-level value as an {index} and $n \geq 0$ arguments, which are other external terms.
 To better motivate this choice, we can give a purely syntactic desugaring of a Python-like syntax with labels to these forms, shown in Figure \ref{desugaring}.
-It is instructive to rewrite lines 27-28 of Figure \ref{record-theory} using these desugarings:% In @\texttt{lang}, desugarings can be user-defined
+It is instructive to rewrite lines 27-28 of Figure \ref{record-theory} using these desugarings:% In \texttt{typy}, desugarings can be user-defined
 %\noindent
@@
 In the core calculus, this is one of three functions, called \FF{iana}, \FF{isyn} and \FF{esyn} in the grammar.\footnote{This means that any particular type constructor only supports a single introduction and elimination desugaring.
-This is a minor inconvenience in some cases that is resolved in @\texttt{lang} by the use of methods.} We see how this is done with the rules for the active typing judgements, given in Figure \ref{atj}.
+This is a minor inconvenience in some cases that is resolved in \texttt{typy} by the use of methods.} We see how this is done with the rules for the active typing judgements, given in Figure \ref{atj}.
 %
@@
 %
-%Compiling a program consists of first \emph{kind checking} it (typechecking the type-level terms, which has no analog in @\texttt{lang} because the type-level language is Python; discussed further below), then typechecking the external term and {translating} it to a term, $\iota$, in the {typed internal language}.
+%Compiling a program consists of first \emph{kind checking} it (typechecking the type-level terms, which has no analog in \texttt{typy} because the type-level language is Python; discussed further below), then typechecking the external term and {translating} it to a term, $\iota$, in the {typed internal language}.
 %These correspond to the premises of the \emph{central compilation judgement} $\pkcompiles{\rho}{\iota}$:
@@
 %%\]
-%%his anchors our exposition; we will describe how it is derived (i.e. how to write a compiler for $\lam@\texttt{lang}$) in the following sections. 
+%%his anchors our exposition; we will describe how it is derived (i.e. how to write a compiler for $\lam\texttt{typy}$) in the following sections. 
 %
@@
 %
-%In $\lam@\texttt{lang}$, the internal language (IL) provides partial functions (via the generic fixpoint operator of Plotkin's PCF), simple product and sum types and a base type of integers (to make our example interesting and as a nod toward speed on contemporary machines).
+%In $\lam\texttt{typy}$, the internal language (IL) provides partial functions (via the generic fixpoint operator of Plotkin's PCF), simple product and sum types and a base type of integers (to make our example interesting and as a nod toward speed on contemporary machines).
 In practice, the internal language could be any typed  language with a specification for which type safety and decidability of typechecking have been satisfyingly determined.
@@
 %\subsection{OpenCL as an Active Library}
-%The code in this section uses \lstinlinep{clx}, an example library implementing the semantics of the OpenCL programming language and extending it with some additional useful types, which we will discuss shortly.
-@\texttt{lang} itself has no built-in support for OpenCL.
+%The code in this section uses \lip{clx}, an example library implementing the semantics of the OpenCL programming language and extending it with some additional useful types, which we will discuss shortly.
+\texttt{typy} itself has no built-in support for OpenCL.
 %
@@
 %%\lstinputlisting[commentstyle=\color{mauve}]{listing7.py}
-%%\caption{[\texttt{listing\ref{metaprogramming}.py}] Metaprogramming with @\texttt{lang}, showing how to construct generic functions from abstract syntax trees.}
+%%\caption{[\texttt{listing\ref{metaprogramming}.py}] Metaprogramming with \texttt{typy}, showing how to construct generic functions from abstract syntax trees.}
 %%\label{metaprogramming}
@@
 %%\lstinputlisting[commentstyle=\color{mauve}]{listing7.py}
-%%\caption{[\texttt{listing\ref{metaprogramming}.py}] Metaprogramming with @\texttt{lang}, showing how to construct generic functions from abstract syntax trees.}
+%%\caption{[\texttt{listing\ref{metaprogramming}.py}] Metaprogramming with \texttt{typy}, showing how to construct generic functions from abstract syntax trees.}
 %%\label{metaprogramming}
 %%\end{codelisting}
-%Lines 3-4 introduce \lstinlinep{map}, an @\texttt{lang} function of three arguments that is governed by the \emph{active base} referred to by \lstinlinep{clx.base} and targeting the \emph{active target} referred to by \lstinlinep{clx.opencl}.
+%Lines 3-4 introduce \lip{map}, an \texttt{typy} function of three arguments that is governed by the \emph{active base} referred to by \lip{clx.base} and targeting the \emph{active target} referred to by \lip{clx.opencl}.
 The active target determines which language the function will compile to (here, the OpenCL kernel language) and mediates code generation.
@@
 %
-%Generic functions are represented at compile-time as instances of \lstinlinep{ace.GenericFn} and consist of an abstract syntax tree, an {active base} and an {active target}.
-The purpose of the \emph{decorator} on line 3 is to replace the Python function on lines 4-8 with an @\texttt{lang} generic function having the same syntax tree and the provided active base and active target.
+%Generic functions are represented at compile-time as instances of \lip{ace.GenericFn} and consist of an abstract syntax tree, an {active base} and an {active target}.
+The purpose of the \emph{decorator} on line 3 is to replace the Python function on lines 4-8 with an \texttt{typy} generic function having the same syntax tree and the provided active base and active target.
 %Decorators in Python are simply syntactic sugar for applying the decorator function directly to the function  being decorated \cite{python}.
@@
 %
-%The abstract syntax tree for \lstinlinep{map} is extracted using the Python standard  library packages  \lstinlinep{inspect} (to retrieve its source code) and \lstinlinep{ast} (to parse it into a syntax tree).
-%
-%%\subsection{Metaprogramming in @\texttt{lang}}
-%%Generic functions can be generated directly from ASTs as well, providing @\texttt{lang} with support for  straightforward metaprogramming.
-Listing \ref{metaprogramming} shows how to generate two more generic functions, \lstinlinep{scale} and \lstinlinep{negate}.
-The latter is derived from the former by using a library for manipulating Python syntax trees, \lstinlinep{astx}.
-In particular, the \lstinlinep{specialize} function replaces uses of the second argument of \lstinlinep{scale} with the literal \lstinlinep{-1} (and changes the function's name), leaving a function of one argument.
+%The abstract syntax tree for \lip{map} is extracted using the Python standard  library packages  \lip{inspect} (to retrieve its source code) and \lip{ast} (to parse it into a syntax tree). 
+%
+%%\subsection{Metaprogramming in \texttt{typy}}
+%%Generic functions can be generated directly from ASTs as well, providing \texttt{typy} with support for  straightforward metaprogramming.
+Listing \ref{metaprogramming} shows how to generate two more generic functions, \lip{scale} and \lip{negate}.
+The latter is derived from the former by using a library for manipulating Python syntax trees, \lip{astx}.
+In particular, the \lip{specialize} function replaces uses of the second argument of \lip{scale} with the literal \lip{-1} (and changes the function's name), leaving a function of one argument.
 % 
@@
 %To compile a generic function to a particular \emph{concrete function}, a type must be provided for each argument, and typechecking and translation must then succeed.
-Listing \ref{compscript} shows how to explicitly provide type assignments to \lstinlinep{map} using the subscript operator (implemented using Python's operator overloading mechanism).
+Listing \ref{compscript} shows how to explicitly provide type assignments to \lip{map} using the subscript operator (implemented using Python's operator overloading mechanism).
 We attempt to do so three times in Listing \ref{compscript}.
@@
 The error occurred  because the ordering of the argument types was incorrect.
-We provide a valid ordering on line \ref{compscript}.9 to generate the concrete function \lstinlinep{map_neg_f32}.
-We then provide a different type assignment to generate the concrete function \lstinlinep{map_neg_ci32}.
-%Concrete functions are instances of \lstinlinep{ace.TypedFn}, consisting of an abstract syntax tree annotated with types and translations along with a reference to the original generic function.
+We provide a valid ordering on line \ref{compscript}.9 to generate the concrete function \lip{map_neg_f32}.
+We then provide a different type assignment to generate the concrete function \lip{map_neg_ci32}.
+%Concrete functions are instances of \lip{ace.TypedFn}, consisting of an abstract syntax tree annotated with types and translations along with a reference to the original generic function.
 %The typing and translation process is mediated by the logic in the active base, types and target that have been provided, as we will describe in more detail below.
 %
-%To produce an output file from an @\texttt{lang} ``compilation script'' like \lstinlinep{listing}\texttt{\ref{compscript}}\lstinlinep{.py}, the command \lstinlinep{acec} can be invoked from the shell, as shown in Listing \ref{mapc}.
+%To produce an output file from an \texttt{typy} ``compilation script'' like \lip{listing}\texttt{\ref{compscript}}\lip{.py}, the command \lip{acec} can be invoked from the shell, as shown in Listing \ref{mapc}.
 %\subsection{Types}
@@
 %\end{codelisting}
-%%Lines \ref{compscript}.3-\ref{compscript}.5 construct the types assigned to the arguments of \lstinlinep{map} on lines \ref{compscript}.7-\ref{compscript}.10.
-In @\texttt{lang}, types are themselves values that can be manipulated at compile-time.
+%%Lines \ref{compscript}.3-\ref{compscript}.5 construct the types assigned to the arguments of \lip{map} on lines \ref{compscript}.7-\ref{compscript}.10.
+In \texttt{typy}, types are themselves values that can be manipulated at compile-time.
 This stands in contrast to other contemporary languages, where user-defined types (e.g.
 datatypes, classes, structs) are written declaratively at compile-time but cannot be constructed, inspected or passed around programmatically.
-More specifically, types are instances of a Python class that implements the \lstinlinep{ace.ActiveType} interface (see Sec.
+More specifically, types are instances of a Python class that implements the \lip{ace.ActiveType} interface (see Sec.
 \ref{atypes}).
-%%As Python values, types can be assigned to variables when convenient (removing the need for  facilities like \lstinlinep{typedef} in C or \lstinlinep{type} in Haskell).
-Types, like all compile-time objects derived from @\texttt{lang} base classes, do not have visible state and operate in a referentially transparent manner (by constructor memoization, which we do not detail here).% These types are all implemented in the \lstinlinep{clx} library imported on line 1, none are built into @\texttt{lang} itself.
-%
-%The type named \lstinlinep{T1} on line \ref{compscript}.3 corresponds to the OpenCL type \lstinlinep{global float*}: a pointer to a 32-bit floating point number stored in the compute device's global memory (one of four address spaces defined by OpenCL \cite{opencl11}).
-It is constructed by applying \lstinlinep{clx.Ptr}, which is an @\texttt{lang} type constructor corresponding to pointer types, to a value representing the  address space, \lstinlinep{clx.global_}, and the type being pointed to.
-That type, \lstinlinep{clx.float}, is in turn the @\texttt{lang} type corresponding to \lstinlinep{float} in OpenCL (which, unlike C99, is always 32 bits).
-%The \lstinlinep{clx} library contains a full implementation of the OpenCL type system (including behaviors, like promotions, inherited from C99).
-%@\texttt{lang} is \emph{unopinionated} about issues like memory safety and the wisdom of such promotions.
+%%As Python values, types can be assigned to variables when convenient (removing the need for  facilities like \lip{typedef} in C or \lip{type} in Haskell).
+Types, like all compile-time objects derived from \texttt{typy} base classes, do not have visible state and operate in a referentially transparent manner (by constructor memoization, which we do not detail here).% These types are all implemented in the \lip{clx} library imported on line 1, none are built into \texttt{typy} itself.
+%
+%The type named \lip{T1} on line \ref{compscript}.3 corresponds to the OpenCL type \lip{global float*}: a pointer to a 32-bit floating point number stored in the compute device's global memory (one of four address spaces defined by OpenCL \cite{opencl11}).
+It is constructed by applying \lip{clx.Ptr}, which is an \texttt{typy} type constructor corresponding to pointer types, to a value representing the  address space, \lip{clx.global_}, and the type being pointed to.
+That type, \lip{clx.float}, is in turn the \texttt{typy} type corresponding to \lip{float} in OpenCL (which, unlike C99, is always 32 bits).
+%The \lip{clx} library contains a full implementation of the OpenCL type system (including behaviors, like promotions, inherited from C99).
+%\texttt{typy} is \emph{unopinionated} about issues like memory safety and the wisdom of such promotions.
 We will discuss how to implement, as libraries, abstractions that are higher-level than raw pointers in Sec.
-\ref{examples}, but @\texttt{lang} does not prevent users from choosing a low level of abstraction or ``interesting'' semantics if the need arises (e.g.
+\ref{examples}, but \texttt{typy} does not prevent users from choosing a low level of abstraction or ``interesting'' semantics if the need arises (e.g.
 for compatibility with existing libraries; see the discussion in Sec.
@@
 We also note that we are being more verbose than necessary for the sake of pedagogy.
-The \lstinlinep{clx} library includes more concise shorthand for OpenCL's types: \lstinlinep{T1} is equal to \lstinlinep{clx.gp(clx.f32)}.
-%Similarly, the decorators in Listings \ref{map} and \ref{metaprogramming} could have been written \lstinlinep{clx.cl_fn}.\todo{move this back there probably}
-%
-%The type \lstinlinep{T2} on line \ref{compscript}.4 is a pointer to a \emph{complex integer} in global memory.
+The \lip{clx} library includes more concise shorthand for OpenCL's types: \lip{T1} is equal to \lip{clx.gp(clx.f32)}.
+%Similarly, the decorators in Listings \ref{map} and \ref{metaprogramming} could have been written \lip{clx.cl_fn}.\todo{move this back there probably}
+%
+%The type \lip{T2} on line \ref{compscript}.4 is a pointer to a \emph{complex integer} in global memory.
 It does not correspond direrctly to a type in OpenCL, because OpenCL does not include primitive support for complex numbers.
-Instead, it uses an active type constructor \lstinlinep{clx.Cplx}, which includes the necessary logic for typechecking operations on complex numbers and translating them to OpenCL (Sec.
+Instead, it uses an active type constructor \lip{clx.Cplx}, which includes the necessary logic for typechecking operations on complex numbers and translating them to OpenCL (Sec.
 \ref{atypes}).
-This constructor is parameterized by the numeric type that should be used for the real and imaginary parts, here \lstinlinep{clx.int}, which corresponds to 32-bit OpenCL integers.
+This constructor is parameterized by the numeric type that should be used for the real and imaginary parts, here \lip{clx.int}, which corresponds to 32-bit OpenCL integers.
 Arithmetic operations with other complex numbers, as well as with plain numeric types (treated as if their imaginary part was zero), are supported.
-When targeting OpenCL, @\texttt{lang} expressions assigned type \lstinlinep{clx.Cplx(clx.int)} are compiled to OpenCL expressions of type \lstinlinep{int2}, a  \emph{vector type} of two 32-bit integers (a type that itself is not inherited from C99).
+When targeting OpenCL, \texttt{typy} expressions assigned type \lip{clx.Cplx(clx.int)} are compiled to OpenCL expressions of type \lip{int2}, a  \emph{vector type} of two 32-bit integers (a type that itself is not inherited from C99).
 This can be observed in several places on lines \ref{mapout}.14-\ref{mapout}.21.
-This choice is merely an implementation detail that can be kept private to \lstinlinep{clx}, however.
-An @\texttt{lang} value of type \lstinlinep{clx.int2} (that is, an actual OpenCL vector) \emph{cannot} be used when a \lstinlinep{clx.Cplx(clx.int)} is expected (and attempting to do so will result in a static type error): \lstinlinep{clx.Cplx} truly extends the type system, it is not a type alias.
-%
-%The type \lstinlinep{TF} on line \ref{compscript}.5 is extracted from the generic function \lstinlinep{negate} constructed in Listing \ref{metaprogramming}.
+This choice is merely an implementation detail that can be kept private to \lip{clx}, however.
+An \texttt{typy} value of type \lip{clx.int2} (that is, an actual OpenCL vector) \emph{cannot} be used when a \lip{clx.Cplx(clx.int)} is expected (and attempting to do so will result in a static type error): \lip{clx.Cplx} truly extends the type system, it is not a type alias.
+%
+%The type \lip{TF} on line \ref{compscript}.5 is extracted from the generic function \lip{negate} constructed in Listing \ref{metaprogramming}.
 Generic functions, according to Sec.
-\ref{genfn}, have not yet had a type assigned to them, so it may seem perplexing that we are nevertheless assigning a type to \lstinlinep{negate}.
-Although a conventional arrow type cannot be assigned to \lstinlinep{negate}, we can give it a \emph{singleton type}: a type that simply means ``this expression is the \emph{particular} generic function \lstinlinep{negate}''.
-This type could also have been explicitly written as \lstinlinep{ace.GenericFnType(listing2.negate)}.
-During typechecking and translation of \lstinlinep{map_neg_f32} and \lstinlinep{map_neg_ci32}, the call to \lstinlinep{f} on line \ref{map}.6 uses the type of the provided argument to compile the generic function that inhabits the singleton type of \lstinlinep{f} (\lstinlinep{negate} in both of these cases) to a concrete function.
-This is why there are two versions of \lstinlinep{negate} in the output in Listing \ref{mapout}.
-In other words, types \emph{propagate} into generic functions -- we didn't need to compile \lstinlinep{negate} explicitly.
-This also explains the error printed on line \ref{mapc}.3-\ref{mapc}.4: when this type was inadvertently assigned to the first argument \lstinlinep{input}, the indexing operation on line \ref{map}.6 resulted in an error.
-A generic function can only be \emph{statically} indexed by a list of types to turn it into a concrete function, not \emph{dynamically} indexed with a value of type \lstinlinep{clx.size_t} (the return type of the OpenCL primitive function \lstinlinep{get_global_id}).
+\ref{genfn}, have not yet had a type assigned to them, so it may seem perplexing that we are nevertheless assigning a type to \lip{negate}.
+Although a conventional arrow type cannot be assigned to \lip{negate}, we can give it a \emph{singleton type}: a type that simply means ``this expression is the \emph{particular} generic function \lip{negate}''.
+This type could also have been explicitly written as \lip{ace.GenericFnType(listing2.negate)}.
+During typechecking and translation of \lip{map_neg_f32} and \lip{map_neg_ci32}, the call to \lip{f} on line \ref{map}.6 uses the type of the provided argument to compile the generic function that inhabits the singleton type of \lip{f} (\lip{negate} in both of these cases) to a concrete function.
+This is why there are two versions of \lip{negate} in the output in Listing \ref{mapout}.
+In other words, types \emph{propagate} into generic functions -- we didn't need to compile \lip{negate} explicitly.
+This also explains the error printed on line \ref{mapc}.3-\ref{mapc}.4: when this type was inadvertently assigned to the first argument \lip{input}, the indexing operation on line \ref{map}.6 resulted in an error.
+A generic function can only be \emph{statically} indexed by a list of types to turn it into a concrete function, not \emph{dynamically} indexed with a value of type \lip{clx.size_t} (the return type of the OpenCL primitive function \lip{get_global_id}).
 %
@@
 Concrete functions, on the other hand, can be given both a singleton type and a true function type.
-For example, \lstinlinep{listing2.negate[[clx.int]]} could be given type \lstinlinep{ace.Arrow(clx.int, clx.int)}.
-The base determines how to convert the @\texttt{lang} arrow type to an arrow type in the target language (e.g.
+For example, \lip{listing2.negate[[clx.int]]} could be given type \lip{ace.Arrow(clx.int, clx.int)}.
+The base determines how to convert the \texttt{typy} arrow type to an arrow type in the target language (e.g.
 a function pointer for C99, or an integer that indexes into a jump table constructed from knowledge of available functions of the appropriate type in OpenCL).
@@
 %Type assignment to generic functions is similar in some ways to template specialization in C++.
-In effect, both a template header and type parameters at call sites are being generated automatically by @\texttt{lang}.
+In effect, both a template header and type parameters at call sites are being generated automatically by \texttt{typy}.
 This simplifies a sophisticated feature of C++ and enables its use with other targets like OpenCL.
@@
 %%\end{codelisting}
-%%On line 5 in the generic \lstinlinep{map} function in Listing \ref{map}, the variable \lstinlinep{gid} is initialized with the result of calling the OpenCL primitive \lstinlinep{get_global_id}.
-The type for \lstinlinep{gid} is never given explicitly.
-This is a simple case of @\texttt{lang}'s {\em within-function type resolution} strategy (we hesitate to call it \emph{type inference} because it does not, strictly speaking, take a constraint-solving approach).
-In this case, the type of \lstinlinep{gid} will resolve to \lstinlinep{size_t} because that is the return type of \lstinlinep{get_global_id} (as defined in the OpenCL specification, which the \lstinlinep{ace.OpenCL} module follows).
+%%On line 5 in the generic \lip{map} function in Listing \ref{map}, the variable \lip{gid} is initialized with the result of calling the OpenCL primitive \lip{get_global_id}.
+The type for \lip{gid} is never given explicitly.
+This is a simple case of \texttt{typy}'s {\em within-function type resolution} strategy (we hesitate to call it \emph{type inference} because it does not, strictly speaking, take a constraint-solving approach).
+In this case, the type of \lip{gid} will resolve to \lip{size_t} because that is the return type of \lip{get_global_id} (as defined in the OpenCL specification, which the \lip{ace.OpenCL} module follows).
 The result can be observed on Lines 11 and 24 in Listing \ref{mapout}.
 %%
-%%Inference is not restricted within single assignments, as in the \lstinlinep{map} example, however.
+%%Inference is not restricted within single assignments, as in the \lip{map} example, however.
 Multiple assignments to the same identifier with values of differing types, or multiple return statements, can be combined if the types in each case are compatible with one another (e.g.
 by a subtyping relation or an implicit coercion).
-In Listing \ref{inference}, the \lstinlinep{threshold_scale} function assigns different values to \lstinlinep{y} in each branch of the conditional.
-In the first branch, the value \lstinlinep{0} is an \lstinlinep{int} literal.
-However, in the second branch of the loop, the type depends on the types of both arguments, \lstinlinep{x} and \lstinlinep{scale}.
+In Listing \ref{inference}, the \lip{threshold_scale} function assigns different values to \lip{y} in each branch of the conditional.
+In the first branch, the value \lip{0} is an \lip{int} literal.
+However, in the second branch of the loop, the type depends on the types of both arguments, \lip{x} and \lip{scale}.
 We show two choices for these types on Lines 11 and 12.
-Type inference correctly combines these two types according to OpenCL's C99-derived rules governing numeric types (defined by the user in the \lstinlinep{OpenCL} module, as we will describe in Section \ref{att}).
+Type inference correctly combines these two types according to OpenCL's C99-derived rules governing numeric types (defined by the user in the \lip{OpenCL} module, as we will describe in Section \ref{att}).
 We can verify this programmatically on Lines 12 and 13.
-Note that this example would also work correctly if the assignments to \lstinlinep{y} were replaced with \lstinlinep{return} statements (in other words, the return value of a function is treated as an assignable for the purpose of type inference).
+Note that this example would also work correctly if the assignments to \lip{y} were replaced with \lip{return} statements (in other words, the return value of a function is treated as an assignable for the purpose of type inference).
 %
@@
 A more recent trend is to generate and compile code without leaving Python, using a Python wrapper around a compiler.
-For example, \lstinlinep{weave} works with C and C++, and \lstinlinep{pycuda} and \lstinlinep{pyopencl} work with CUDA and OpenCL, respectively \cite{klockner2011pycuda}.
+For example, \lip{weave} works with C and C++, and \lip{pycuda} and \lip{pyopencl} work with CUDA and OpenCL, respectively \cite{klockner2011pycuda}.
 %\begin{codelisting}
@@
 %The OpenCL language was designed for this workflow, exposing a retargetable compiler and data management  routines as an API, called the \emph{host API} \cite{opencl11}.
-The \lstinlinep{pyopencl} library exposes this API to Python and simplifies interoperation with \lstinlinep{numpy}, a popular package for manipulating contiguously-allocated numeric arrays in Python \cite{klockner2011pycuda}.
-%
-%@\texttt{lang} supports a refinement to this workflow, as an alternative to the \lstinlinep{acec} compiler described above, for targets that have wrappers like this available, including \lstinlinep{clx.opencl}.
-Listing \ref{py} shows an example of this workflow where the user chooses a compute device (line \ref{py}.3),  constructs a \lstinlinep{numpy} array (line \ref{py}.5), transfers it to the device (line \ref{py}.6), allocates an empty equal-sized buffer for the result of the computation (line \ref{py}.7), launches the {generic} kernel \lstinlinep{map} from Listing \ref{map} with these device arrays as well as the function \lstinlinep{negate} from Listing \ref{metaprogramming} (line \ref{py}.9) choosing a number of threads equal to the number of elements in the input array (line \ref{py}.10), and transfers the result back into main memory to check that the device computed the expected result (line \ref{py}.12).
+The \lip{pyopencl} library exposes this API to Python and simplifies interoperation with \lip{numpy}, a popular package for manipulating contiguously-allocated numeric arrays in Python \cite{klockner2011pycuda}.
+%
+%\texttt{typy} supports a refinement to this workflow, as an alternative to the \lip{acec} compiler described above, for targets that have wrappers like this available, including \lip{clx.opencl}.
+Listing \ref{py} shows an example of this workflow where the user chooses a compute device (line \ref{py}.3),  constructs a \lip{numpy} array (line \ref{py}.5), transfers it to the device (line \ref{py}.6), allocates an empty equal-sized buffer for the result of the computation (line \ref{py}.7), launches the {generic} kernel \lip{map} from Listing \ref{map} with these device arrays as well as the function \lip{negate} from Listing \ref{metaprogramming} (line \ref{py}.9) choosing a number of threads equal to the number of elements in the input array (line \ref{py}.10), and transfers the result back into main memory to check that the device computed the expected result (line \ref{py}.12).
 %
 %For developers experienced with the usual OpenCL or CUDA workflow, the fact that this can be accomplished in a total of 6 statements may be surprising.
-This simplicity is made possible by @\texttt{lang}'s implicit tracking of types throughout the code.
-First, \lstinlinep{numpy} keeps track of the type, shape and order of its arrays.
-The type of \lstinlinep{input}, for example, is \lstinlinep{numpy.float64} by default, its shape is \lstinlinep{(1024,1024)} and its order is row-major by default.
-The \lstinlinep{pyopencl} library, which the \lstinlinep{clx} mechanism is built upon, uses this metadata to automatically call the underlying OpenCL host API function for transferring byte arrays to the device without requiring the user to calculate the size.
-The \lstinlinep{clx} wrapper further retains this metadata in \lstinlinep{d_input} and \lstinlinep{d_output}, the Python wrappers around the allocated device arrays.
-The @\texttt{lang} active type of these wrappers is an instance of \lstinlinep{clx.NPArray} parameterized by this metadata.
+This simplicity is made possible by \texttt{typy}'s implicit tracking of types throughout the code.
+First, \lip{numpy} keeps track of the type, shape and order of its arrays.
+The type of \lip{input}, for example, is \lip{numpy.float64} by default, its shape is \lip{(1024,1024)} and its order is row-major by default.
+The \lip{pyopencl} library, which the \lip{clx} mechanism is built upon, uses this metadata to automatically call the underlying OpenCL host API function for transferring byte arrays to the device without requiring the user to calculate the size.
+The \lip{clx} wrapper further retains this metadata in \lip{d_input} and \lip{d_output}, the Python wrappers around the allocated device arrays.
+The \texttt{typy} active type of these wrappers is an instance of \lip{clx.NPArray} parameterized by this metadata.
 This type knows how to typecheck and translate operations, like indexing with a multi-dimensional thread index, automatically.
 %
-%By extracting the types from the arguments, we can call the generic function \lstinlinep{map} without first requiring an explicit type assignment, like we needed when using \lstinlinep{acec} above.
-In other words, dynamic types and other metadata can propagate from Python data structures into an @\texttt{lang} generic function as static type information, in the same manner as it propagated \emph{between} generic functions in the previous section.
-In both cases, typechecking and translation of \lstinlinep{map} happens the first time a particular type assignment is encountered and cached for subsequent use.
+%By extracting the types from the arguments, we can call the generic function \lip{map} without first requiring an explicit type assignment, like we needed when using \lip{acec} above.
+In other words, dynamic types and other metadata can propagate from Python data structures into an \texttt{typy} generic function as static type information, in the same manner as it propagated \emph{between} generic functions in the previous section.
+In both cases, typechecking and translation of \lip{map} happens the first time a particular type assignment is encountered and cached for subsequent use.
 When called from Python, the generated OpenCL source code is compiled for the device we selected using the OpenCL retargetable compilation infrastructure, and cached for subsequent calls.
@@
 %The same program written using the OpenCL C API directly is an order of magnitude longer and significantly more difficult to comprehend.
-OpenCL does not support higher-order functions nor is there any way to write \lstinlinep{map} in a type-generic or shape-generic manner.
-If we instead use the \lstinlinep{pyopencl} library and apply the techniques described in \cite{klockner2011pycuda}, the program is still twice as large and less readable than this code.
-Both the \lstinlinep{map} and \lstinlinep{negate} functions must be explicitly specialized with the appropriate types using string manipulation techniques, and custom shapes and orders can be awkward to handle.
+OpenCL does not support higher-order functions nor is there any way to write \lip{map} in a type-generic or shape-generic manner.
+If we instead use the \lip{pyopencl} library and apply the techniques described in \cite{klockner2011pycuda}, the program is still twice as large and less readable than this code.
+Both the \lip{map} and \lip{negate} functions must be explicitly specialized with the appropriate types using string manipulation techniques, and custom shapes and orders can be awkward to handle.
 Higher order functions are still not available, and must also be simulated by string manipulation.
-That approach also does not permit the use any of the language extensions that @\texttt{lang} enables (beyond the useful type for \lstinlinep{numpy} arrays just described; see Sec.
+That approach also does not permit the use any of the language extensions that \texttt{typy} enables (beyond the useful type for \lip{numpy} arrays just described; see Sec.
 \ref{examples} for more interesting possibilities).
 %
-%%Not shown are several additional conveniences, such as delegated kernel sizing and \lstinlinep{In} and \lstinlinep{Out} constructs that can reduce the size and improve the clarity of this code further; due to a lack of space, the reader is referred to the language documentation.
+%%Not shown are several additional conveniences, such as delegated kernel sizing and \lip{In} and \lip{Out} constructs that can reduce the size and improve the clarity of this code further; due to a lack of space, the reader is referred to the language documentation.
 
@@
 New types are not merely aliases, nor must their rules be admissible in some base type system (e.g.
-\lstinlinep{string_in} would be difficult to define in Scala, particularly as directly as we can here).
-Note that in @\texttt{lang}, macros can be seen as a mode of use of the term constructor \verb|Call|, as we can give the macro a singleton type (like Python modules, discussed above) that performs the desired rewriting in \verb|ana_Call| and \verb|syn_Call|.
+\lip{string_in} would be difficult to define in Scala, particularly as directly as we can here).
+Note that in \texttt{typy}, macros can be seen as a mode of use of the term constructor \verb|Call|, as we can give the macro a singleton type (like Python modules, discussed above) that performs the desired rewriting in \verb|ana_Call| and \verb|syn_Call|.
 
 %Generic functions are a novel strategy for {\it function polymorphism} -- defining functions that operate over more than a single type.
-In @\texttt{lang}, generic functions are implicitly polymorphic and can be called with arguments of {\it any type that supports the operations used by the function}.
+In \texttt{typy}, generic functions are implicitly polymorphic and can be called with arguments of {\it any type that supports the operations used by the function}.
 This is related to structural polymorphism \cite{malayeri2009structural}.
@@
 
-Although we only touched on the details here, we have conducted a substantial case study using @\texttt{lang}: an implementation of the entirety of the OpenCL type system (a variant of C99), as well as several extensions to it, as a library.
+Although we only touched on the details here, we have conducted a substantial case study using \texttt{typy}: an implementation of the entirety of the OpenCL type system (a variant of C99), as well as several extensions to it, as a library.
 Our operations translate to Python's lower-level FFI with OpenCL but guarantee type safety at the interface between languages.
@@
 We believe that by enforcing strict abstraction barriers between extensions, we can approach this goal (and a manuscript is in preparation).
-Bootstrapping the @\texttt{lang} compiler would be an interesting direction to explore, to avoid the awkwardness of relying on a loose dynamically typed language to implement static type systems.
+Bootstrapping the \texttt{typy} compiler would be an interesting direction to explore, to avoid the awkwardness of relying on a loose dynamically typed language to implement static type systems.
 % Different dialects can be difficult to combine manually, as we discussed in the introduction.
@@
 %\subsection{Compositional Reasoning}\label{safety}
-%Every statement and expression in an @\texttt{lang} function is governed by exactly one active type, determined by the dispatch protocol, or by the single active base associated with the @\texttt{lang} function.
+%Every statement and expression in an \texttt{typy} function is governed by exactly one active type, determined by the dispatch protocol, or by the single active base associated with the \texttt{typy} function.
 The representational consistency check ensures that compilation is successful without requiring that types that abstract over other types (e.g.
 container types) know about their internal implementation details.
-Together, this permits compositional reasoning about the semantics of @\texttt{lang} functions even in the presence of many extensions.
+Together, this permits compositional reasoning about the semantics of \texttt{typy} functions even in the presence of many extensions.
 This stands on contrast to many prior approaches to extensibility, where extensions could insert themselves into the semantics in conflicting ways, making the order of imports matter (see Sec.
@@
 %\begin{enumerate}
-%\item We address the ``chicken-and-egg'' problem by designing @\texttt{lang}, an extensible language  embedded within Python \cite{Politz:2013:PFM:2509136.2509536,python} entirely as a library.
-The top-level of an @\texttt{lang} file is a \emph{compilation script} written in Python and Python serves as @\texttt{lang}'s \emph{type-level language}, as we will discuss.
-@\texttt{lang} and Python thus share a common syntax and package system, so users of @\texttt{lang} can directly leverage established tools, infrastructure and coding standards.
-Functions marked by a decorator as under the control of @\texttt{lang} are, unlike Python functions, \emph{statically} typechecked.
+%\item We address the ``chicken-and-egg'' problem by designing \texttt{typy}, an extensible language  embedded within Python \cite{Politz:2013:PFM:2509136.2509536,python} entirely as a library.
+The top-level of an \texttt{typy} file is a \emph{compilation script} written in Python and Python serves as \texttt{typy}'s \emph{type-level language}, as we will discuss.
+\texttt{typy} and Python thus share a common syntax and package system, so users of \texttt{typy} can directly leverage established tools, infrastructure and coding standards.
+Functions marked by a decorator as under the control of \texttt{typy} are, unlike Python functions, \emph{statically} typechecked.
 This makes it possible to rule out many potential issues ahead of execution, and thus  avoid  costly dynamic checks.
 %The semantics of  define run-time behavior.
-%\item  % Instead, @\texttt{lang} has a statically-typed semantics that can be extended by users from within libraries. 
-%There is \emph{no particular type system built into @\texttt{lang}}.
-Instead, a type is any  object implementing the \lstinlinep{ace.Type} interface constructed by the compilation script.
+%\item  % Instead, \texttt{typy} has a statically-typed semantics that can be extended by users from within libraries. 
+%There is \emph{no particular type system built into \texttt{typy}}.
+Instead, a type is any  object implementing the \lip{ace.Type} interface constructed by the compilation script.
 We call these \emph{active types}.
 When statically assigning a type to a compound form, e.g.
-\lstinlinep{e.x}, the @\texttt{lang} {compiler} delegates to the type of a subexpression, e.g.
-to the type of \lstinlinep{e}, to determine how to assign a type to the expression as a whole.
+\lip{e.x}, the \texttt{typy} {compiler} delegates to the type of a subexpression, e.g.
+to the type of \lip{e}, to determine how to assign a type to the expression as a whole.
 As we will show, this {type-directed protocol}, \emph{active typing}, permits the expression of a rich variety of type systems as libraries.
@@
 Base cases, e.g.
-variables and certain statement forms, are handled on a {per-function basis} by a \emph{base semantics}, also a compile-time object implementing an interface, \lstinlinep{ace.Base}.
+variables and certain statement forms, are handled on a {per-function basis} by a \emph{base semantics}, also a compile-time object implementing an interface, \lip{ace.Base}.
 %\end{enumerate}
@@
 %
-%The dynamic behavior of an @\texttt{lang} function is determined by translation to a \emph{target language}.
+%The dynamic behavior of an \texttt{typy} function is determined by translation to a \emph{target language}.
 We begin by  targeting Python, then discuss targeting OpenCL and CUDA, lower-level languages used to program many-core processors (e.g.
@@
 We call this phase \emph{active translation}.
-A \emph{target} mediates this process and is also a user-defined compile-time object implementing an interface, \lstinlinep{ace.Target}.
+A \emph{target} mediates this process and is also a user-defined compile-time object implementing an interface, \lip{ace.Target}.
 %%One base or type can support multiple targets. 
-%When targeting a typed language, care must be taken to ensure that well-typed @\texttt{lang} expressions have well-typed translations, so we integrate a technique for compositionally reasoning about compiler correctness (developed for the TIL compiler for Standard ML \cite{tarditi+:til-OLD}) into the language.
+%When targeting a typed language, care must be taken to ensure that well-typed \texttt{typy} expressions have well-typed translations, so we integrate a technique for compositionally reasoning about compiler correctness (developed for the TIL compiler for Standard ML \cite{tarditi+:til-OLD}) into the language.
 %
@@
 Dynamically-typed languages, on the other hand, avoid most of this overhead by relying on support from the run-time system.
-@\texttt{lang} was first conceived to explore the question: \emph{does conciseness require run-time mechanisms, or can one develop a statically-typed language with the same usability profile?}
+\texttt{typy} was first conceived to explore the question: \emph{does conciseness require run-time mechanisms, or can one develop a statically-typed language with the same usability profile?}
 %
@@
 %Rather than designing a new syntax, or modifying the syntax of an existing language, we chose to utilize, \emph{without modification}, the syntax of an existing language, Python.
-This choice was not arbitrary, but rather a key means by which @\texttt{lang} satisfies extrinsic design criteria related to familiarity and tool support.
+This choice was not arbitrary, but rather a key means by which \texttt{typy} satisfies extrinsic design criteria related to familiarity and tool support.
 Researchers often dismiss the importance of syntax.
@@
 We have developed a simplified, type-theoretic formalism of active typechecking and translation, where we prove the safety properties we have outlined here, as well as several additional ones that are only possible to achieve using a typed metalanguage (currently in submission at ESOP 2014).
-The work described here  extends that mechanism in several directions: @\texttt{lang} supports a rich syntax, makes syntax trees available to extensions, supports a pluggable backend and per-function base semantics.
+The work described here  extends that mechanism in several directions: \texttt{typy} supports a rich syntax, makes syntax trees available to extensions, supports a pluggable backend and per-function base semantics.
 It is implemented in an existing language and supports a wider range of practical extensions.
@@
 
+\section{Implementation}
+We are actively developing the \verb|typy| library and its surrounding ecosystem as a free open source project. The \lip{typy} website can be found at \url{http://www.cs.cmu.edu/~comar/typy/}.
+
+\section{Conclusion}This appears to be a widespread problem: programmers and development teams often cannot use  the constructs they might prefer because they are only available in specialized dialects they cannot  adopt \cite{Meyerovich:2012:SDR:2414721.2414724,Meyerovich:2013:EAP:2509136.2509515}.\todo{finish this} %
+
 % We recommend abbrvnat bibliography style.
-\newpage
+%\newpage
 \bibliographystyle{abbrv}
@@
 %\section{Source Code Emission}
-%Portions of the \lstinlinep{Target} class related to source code emission are shown.
-If compilation occurred externally, the \lstinlinep{acec} library asks the targets used by the top-level typed functions to generate one or more files by deciding on a name and file extension on the basis of the name of the compilation script, and emitting code and \emph{snippets}, which are simply strings corresponding to functions, type declarations, imports and other top-level entities in the target language, each inserted at a \emph{location}.
+%Portions of the \lip{Target} class related to source code emission are shown.
+If compilation occurred externally, the \lip{acec} library asks the targets used by the top-level typed functions to generate one or more files by deciding on a name and file extension on the basis of the name of the compilation script, and emitting code and \emph{snippets}, which are simply strings corresponding to functions, type declarations, imports and other top-level entities in the target language, each inserted at a \emph{location}.
 Each file being generated is a location, and locations might also be nested, depending on the language (e.g.
-the import block may be a separate location when generating Java, or the body of the \lstinlinep{<head>} tag when generating HTML+CSS+Javascript).
+the import block may be a separate location when generating Java, or the body of the \lip{<head>} tag when generating HTML+CSS+Javascript).
 Snippets are only added once to a particular location (so imports are not duplicated, for example).
@@
 %
-%If \lstinlinep{c} is a complex number, then \lstinlinep{c.ni} and \lstinlinep{c.i} are its non-imaginary and imaginary components, respectively.
-These expressions are of the form \lstinlinep{Attribute}, so the typechecker calls \lstinlinep{type_Attribute} (line \ref{cplx}.7).
-%This method receives the compilation context, \lstinlinep{context}, and the abstract syntax tree of the expression, \lstinlinep{node} and must return a type assignment for the node, or raise an \lstinlinep{ace.TypeError} if there is an error.
-In this case, a type assignment is possible if the attribute name is either \lstinlinep{"ni"} or \lstinlinep{"i"}, and an error is raised otherwise (lines \ref{cplx}.8-\ref{cplx}.10).
+%If \lip{c} is a complex number, then \lip{c.ni} and \lip{c.i} are its non-imaginary and imaginary components, respectively.
+These expressions are of the form \lip{Attribute}, so the typechecker calls \lip{type_Attribute} (line \ref{cplx}.7).
+%This method receives the compilation context, \lip{context}, and the abstract syntax tree of the expression, \lip{node} and must return a type assignment for the node, or raise an \lip{ace.TypeError} if there is an error.
+In this case, a type assignment is possible if the attribute name is either \lip{"ni"} or \lip{"i"}, and an error is raised otherwise (lines \ref{cplx}.8-\ref{cplx}.10).
 We note that error messages are an important and sometimes overlooked facet of {ease-of-use} \cite{marceau2011measuring}.
 A common frustration with using general-purpose abstraction mechanisms to encode an abstraction is that they can produce  verbose and cryptic error messages that reflect the implementation details instead of the semantics.
-@\texttt{lang} supports custom error messages.%Active types permit direct encoding of semantics and customization of error messages.
+\texttt{typy} supports custom error messages.%Active types permit direct encoding of semantics and customization of error messages.
 %
@@
 Arithmetic operations are usually symmetric, so the dispatch protocol checks the types of both subexpressions for support.
-To ensure that the semantics remain deterministic in the case that both types support the binary operation, @\texttt{lang} asks the left first (via \lstinlinep{type_BinOp_left}), asking the right (via \lstinlinep{type_BinOp_right}) only if the left indicates an error.
-In either position, our implementation begins by recursively assigning a type to the other operand in the current context via the \lstinlinep{context.type} method (line \ref{cplx}.24).
-If supported, it applies the C99 rules for arithmetic operations to determine the resulting type (via \lstinlinep{c99_binop_t}, not shown).
-%
-%Finally, a complex number can be constructed inside an @\texttt{lang} function using @\texttt{lang}'s special constructor form: \lstinlinep{[clx.Cplx](3,4)} represents $3+4i$, for example.
+To ensure that the semantics remain deterministic in the case that both types support the binary operation, \texttt{typy} asks the left first (via \lip{type_BinOp_left}), asking the right (via \lip{type_BinOp_right}) only if the left indicates an error.
+In either position, our implementation begins by recursively assigning a type to the other operand in the current context via the \lip{context.type} method (line \ref{cplx}.24).
+If supported, it applies the C99 rules for arithmetic operations to determine the resulting type (via \lip{c99_binop_t}, not shown).
+%
+%Finally, a complex number can be constructed inside an \texttt{typy} function using \texttt{typy}'s special constructor form: \lip{[clx.Cplx](3,4)} represents $3+4i$, for example.
 The term within the braces is evaluated at \emph{compile-time}.
-Because \lstinlinep{clx.Cplx} evaluates not to an active type, but to a class, this form is assigned a type by handing control to the class object via the \emph{class method} \lstinlinep{type_New}.
+Because \lip{clx.Cplx} evaluates not to an active type, but to a class, this form is assigned a type by handing control to the class object via the \emph{class method} \lip{type_New}.
 It operates as expected, extracting the types of the two arguments to construct an appropriate complex number type (lines \ref{cplx}.50-\ref{cplx}.57), raising a type error if the arguments cannot be promoted to a common type according to the rules of C99 or if two arguments were not provided.% (an exercise for the reader: modify this method to also allow a single argument for when the imaginary part is 0).
 %
-%As seen in Listing \ref{mapout}, we are implementing complex numbers internally using OpenCL vector types, like \lstinlinep{int2}.
-Let us look first at \lstinlinep{trans_New} on lines \ref{cplx}.54-\ref{cplx}.60, where new complex numbers are translated to vector literals by invoking \lstinlinep{target.VecLit}.
+%As seen in Listing \ref{mapout}, we are implementing complex numbers internally using OpenCL vector types, like \lip{int2}.
+Let us look first at \lip{trans_New} on lines \ref{cplx}.54-\ref{cplx}.60, where new complex numbers are translated to vector literals by invoking \lip{target.VecLit}.
 This will ultimately  generate the necessary OpenCL code, as a string, to complete compilation (these strings are not directly manipulated by extensions, however, to avoid problems with, e.g.
@@
 For it to be possible to reason compositionally about the correctness of compilation, all complex numbers must translate to terms in the target language that have a consistent target type.
-The \lstinlinep{trans_type} method of the \lstinlinep{ace.ActiveType} associates a type in the target language, here a vector type like \lstinlinep{int2}, with the active type.
-@\texttt{lang} supports a mode where this \emph{representational consistency} is dynamically checked during compilation (requiring that the active target know how to assign types to terms in the target language, which can be done  for our OpenCL target as of this writing).
+The \lip{trans_type} method of the \lip{ace.ActiveType} associates a type in the target language, here a vector type like \lip{int2}, with the active type.
+\texttt{typy} supports a mode where this \emph{representational consistency} is dynamically checked during compilation (requiring that the active target know how to assign types to terms in the target language, which can be done  for our OpenCL target as of this writing).
 %
 %The translation methods for attributes (line \ref{cplx}.12) and binary operations  (line \ref{cplx}.31) proceed in a straightforward manner.
-The context provides a method, \lstinlinep{trans}, for recursively determining the translation of subexpressions as needed.
+The context provides a method, \lip{trans}, for recursively determining the translation of subexpressions as needed.
 Of note is that the translation methods can assume that typechecking succeeded.
-For example, the implementation of \lstinlinep{trans_Attribute} assumes that if \lstinlinep{node.attr} is not \lstinlinep{'ni'} then it must have been \lstinlinep{'i'} on line \ref{cplx}.14, consistent with the implementation of \lstinlinep{type_Attribute} above it.
+For example, the implementation of \lip{trans_Attribute} assumes that if \lip{node.attr} is not \lip{'ni'} then it must have been \lip{'i'} on line \ref{cplx}.14, consistent with the implementation of \lip{type_Attribute} above it.
 Typechecking and translation are separated into two methods to emphasize that typechecking is not target-dependent, and to allow for more advanced uses, like type refinements and hypothetical typing judgements, that we do not describe here.
@@
 %Thus far, we have focused mainly on the OpenCL target and shown examples of fairly low-level active types: those that implement OpenCL's primitives (e.g.
-\lstinlinep{clx.Ptr}), extend them in simple but convenient ways (e.g.
-\lstinlinep{clx.Cplx}) and those that make interactive execution across language boundaries safe and convenient (e.g.
-\lstinlinep{clx.NPArray}).
-%@\texttt{lang} was first conceived to answer the question: \emph{can we build a statically-typed language with the semantics of a C but the syntax and ease-of-use of a Python?} We submit that the work described above has answered this question in the affirmative.
-%
-%But @\texttt{lang} has proven useful for more than low-level tasks like programming a GPU with OpenCL.
+\lip{clx.Ptr}), extend them in simple but convenient ways (e.g.
+\lip{clx.Cplx}) and those that make interactive execution across language boundaries safe and convenient (e.g.
+\lip{clx.NPArray}).
+%\texttt{typy} was first conceived to answer the question: \emph{can we build a statically-typed language with the semantics of a C but the syntax and ease-of-use of a Python?} We submit that the work described above has answered this question in the affirmative.
+%
+%But \texttt{typy} has proven useful for more than low-level tasks like programming a GPU with OpenCL.
 We now describe several interesting extensions that implement the semantics of primitives drawn from a range of different language paradigms, to justify our claim that these mechanisms are highly expressive.
 %
-%\subsection{Growing a Statically-Typed Python Inside an @\texttt{lang}}
-%@\texttt{lang} comes with a target, base and type implementing Python itself: \lstinlinep{ace.python.python}, \lstinlinep{ace.python.base} and \lstinlinep{ace.python.dyn}.
+%\subsection{Growing a Statically-Typed Python Inside an \texttt{typy}}
+%\texttt{typy} comes with a target, base and type implementing Python itself: \lip{ace.python.python}, \lip{ace.python.base} and \lip{ace.python.dyn}.
 These can be supplemented by additional active types and used as the foundation for writing actively-typed Python functions.
@@
 recursive labeled sum types) together with the Python implementation just described.
-It shows two syntactic conveniences that were not mentioned previously: 1) if no target is provided to \lstinlinep{ace.fn}, the base can provide a default target (here, \lstinlinep{ace.python.python}); 2) a concrete function can be generated immediately by providing a type assignment after \lstinlinep{ace.fn} using braces.
+It shows two syntactic conveniences that were not mentioned previously: 1) if no target is provided to \lip{ace.fn}, the base can provide a default target (here, \lip{ace.python.python}); 2) a concrete function can be generated immediately by providing a type assignment after \lip{ace.fn} using braces.
 Listing \ref{datatypest} generates Listing \ref{datatypes}.
 %
-%Lines \ref{datatypest}.3-\ref{datatypest}.11 define a function that generates a recursive algebraic datatype representing a tree given a name and another @\texttt{lang} type for the data at the leaves.
-This type implemented by the active type family \lstinlinep{fp.Datatype}.
+%Lines \ref{datatypest}.3-\ref{datatypest}.11 define a function that generates a recursive algebraic datatype representing a tree given a name and another \texttt{typy} type for the data at the leaves.
+This type implemented by the active type family \lip{fp.Datatype}.
 A name for the datatype and the case names and types are provided programmatically on lines \ref{datatypest}.3-\ref{datatypest}.8.
@@
 These lines also show two more active type families: units (the type containing just one value), and immutable pairs.
-Line \ref{datatypest}.13 calls this function with the @\texttt{lang} type for dynamic Python values, \lstinlinep{py.dyn}, to generate a type, aliased \lstinlinep{DT}.
+Line \ref{datatypest}.13 calls this function with the \texttt{typy} type for dynamic Python values, \lip{py.dyn}, to generate a type, aliased \lip{DT}.
 This type is implemented using class inheritance when the target is Python, as seen on lines \ref{datatypes}.1-\ref{datatypes}.13.
-For C-like targets, a \lstinlinep{union} type can be used (not shown).
-%
-%The generic function \lstinlinep{depth_gt_2} demonstrates two features.
+For C-like targets, a \lip{union} type can be used (not shown).
+%
+%The generic function \lip{depth_gt_2} demonstrates two features.
 First, the base has been setup to treat the final expression in a top-level branch as the return value of the function, consistent with typical functional languages.
-Second, the \lstinlinep{case} ``method'' on line \ref{datatypest}.17 creatively reuses the syntax for dictionary literals to express a nested pattern matching construct.
+Second, the \lip{case} ``method'' on line \ref{datatypest}.17 creatively reuses the syntax for dictionary literals to express a nested pattern matching construct.
 The patterns to the left of the colons are not treated as expressions (that is, a type and translation is never recursively assigned to them).
@@
 %%Functional constructs have shown promise as the basis for a nested data-parallel parallel programming model.
-Many powerful parallel algorithms can be specified by composing primitives like \lstinlinep{map} and \lstinlinep{reduce} on persistent data structures like lists, trees and records.
+Many powerful parallel algorithms can be specified by composing primitives like \lip{map} and \lip{reduce} on persistent data structures like lists, trees and records.
 Data dependencies are directly encoded in these specifications, and automatic code generation techniques have shown promise in using this information to automatically execute these specifications on concurrent hardware and networks.
@@
 By combining active types and the metaprogramming facilities described in Sec.
-2 to implement optimizations, like fusion, on the typed syntax trees available from concrete functions, these same techniques can be implemented as an @\texttt{lang} library as well.
+2 to implement optimizations, like fusion, on the typed syntax trees available from concrete functions, these same techniques can be implemented as an \texttt{typy} library as well.
 %
 %\subsection{Product Types}
-%Product types (types that group heterogeneous values together) like structs, unions, tuples, records and objects can all be implemented as @\texttt{lang} type families  parameterized by a dictionary mapping field names (indices, in the case of tuples) to types.
+%Product types (types that group heterogeneous values together) like structs, unions, tuples, records and objects can all be implemented as \texttt{typy} type families  parameterized by a dictionary mapping field names (indices, in the case of tuples) to types.
 Each family differs slightly in the semantics of the operations available.
 Structs support mutation, for example, while records do not.
-Object systems typically introduce additional logic for calling methods, binding special variables like \lstinlinep{this}, and accessing information through an inheritance hierarchy.
+Object systems typically introduce additional logic for calling methods, binding special variables like \lip{this}, and accessing information through an inheritance hierarchy.
 Listing \ref{ooclxfppy} shows an example of each of these.
 The prototypic object system delegates to a backing record if a field is not available in the foreground.
-This example also demonstrates cross-compilation to C99, supported for a subset of operations in \lstinlinep{clx}.
-All three abstractions are implemented as simple \lstinlinep{struct}s, as can be seen in Listing \ref{ooclxfpc}.
+This example also demonstrates cross-compilation to C99, supported for a subset of operations in \lip{clx}.
+All three abstractions are implemented as simple \lip{struct}s, as can be seen in Listing \ref{ooclxfpc}.
 This example further demonstrates the use of argument annotations to generate concrete functions (line \ref{ooclxfppy}.9), which is a feature only available when using Python 3.3+.
@@
 For example, in Charm++, dynamic dispatch involves message passing over a dynamically load-balanced network \cite{kale2009charm++}.
-While Charm++ is a sophisticated system, and we do not anticipate that implementing it as an @\texttt{lang} extension would be easy, it is possible to do so by targeting a system like Adaptive MPI, which exposes the Charm++ runtime system \cite{kale2009charm++}.
+While Charm++ is a sophisticated system, and we do not anticipate that implementing it as an \texttt{typy} extension would be easy, it is possible to do so by targeting a system like Adaptive MPI, which exposes the Charm++ runtime system \cite{kale2009charm++}.
 %Erlang and other message-passing and actor-based systems can also be considered in this manner -- processes can be thought of as objects, channels as methods and messages as calls.
@@
 %%In addition to type annotations, OpenCL normally asks for additional annotations in a number of other situations.
-Users can annotate functions that meet certain requirements  with the \lstinlinep{__kernel} attribute, indicating that they are callable from the host.
-The \lstinlinep{OpenCL} backend can check these and add this annotation automatically.
-Several types (notably, \lstinlinep{double}) and specialized primitive functions require that an OpenCL extension be enabled via a \lstinlinep{#pragma}.
-The OpenCL backend automatically detects many of these cases as well, adding the appropriate \lstinlinep{#pragma} declaration.
-An example of this can be seen in Listing \ref{mapout}, where the use of the \lstinlinep{double} type triggers the insertion of the \lstinlinep{cl_khr_f64} extension.
+Users can annotate functions that meet certain requirements  with the \lip{__kernel} attribute, indicating that they are callable from the host.
+The \lip{OpenCL} backend can check these and add this annotation automatically.
+Several types (notably, \lip{double}) and specialized primitive functions require that an OpenCL extension be enabled via a \lip{#pragma}.
+The OpenCL backend automatically detects many of these cases as well, adding the appropriate \lip{#pragma} declaration.
+An example of this can be seen in Listing \ref{mapout}, where the use of the \lip{double} type triggers the insertion of the \lip{cl_khr_f64} extension.
 %%
@@
 %\subsubsection{Units of Measure}
-%A number of domain-specific type systems can be implemented within @\texttt{lang} as well.
+%A number of domain-specific type systems can be implemented within \texttt{typy} as well.
 For example, prior work has considered tracking units of measure (e.g.
@@
 This cannot easily be implemented using many existing abstraction mechanisms because this information should only be maintained statically to avoid excessive run-time overhead associated with tagging/boxing, and the typechecking logic is reasonably complex.
-The @\texttt{lang} extension mechanism allows this information to be tracked in the type system, but not included during translation, by a mechanism similar to how address space information is tracked with pointers.
-Because @\texttt{lang} extensions can use Python, the needed logic can be implemented directly.
+The \texttt{typy} extension mechanism allows this information to be tracked in the type system, but not included during translation, by a mechanism similar to how address space information is tracked with pointers.
+Because \texttt{typy} extensions can use Python, the needed logic can be implemented directly.
 %

### 2016-09-17 typy-2016/test.py
put extended version into repo

@@ -1,10 +1,0 @@
-from numpy import (ndarray, 
-  array)
-
-def f(x): 
-  """Put stuff here."""
-  {
-  x : int, # x is an integer
-  y : int  # y is an integer
-  } > int
-  return x

### 2016-09-19 typy-2016/CMU-ISR-16-112.tex
wrap up

\documentclass{article}
\usepackage[fullpage]{cmu-titlepage2}
% \usepackage{fullpage}
\usepackage[letterpaper]{geometry}
% \usepackage{times}
% \usepackage{helvet}
% \usepackage{courier}
\title{Programmable Semantic Fragments \\(Extended Version)}
\author{Cyrus Omar \and Jonathan Aldrich}
\date{September 2016}
\trnumber{CMU-ISR-16-112}
\keywords{metaprogramming, bidirectional typechecking, pattern matching, foreign function interfaces}
\support{~}
\abstract{This technical report serves to archive the extended version of the paper titled \emph{Programmable Semantic Fragments}, which appears in the proceedings of the 15th International Conference on Generative Programming: Concepts \& Experience (GPCE 2016).}
\begin{document}
\maketitle

% \tableofcontents
% \listoffigures
% \clearpage
% \section{Test}

\end{document}

### 2016-10-24 typy-2016/CMU-ISR-16-112.tex
added preprint and page number markers to online text

@@
 % \usepackage{courier}
-\title{Programmable Semantic Fragments \\(Extended Version)}
+\title{Programmable Semantic Fragments: The Design and Implementation of \texttt{typy} (Extended Version)}
 \author{Cyrus Omar \and Jonathan Aldrich}
-\date{September 2016}
+\date{October 2016}
 \trnumber{CMU-ISR-16-112}
@@
 \support{~}
-\abstract{This technical report serves to archive the extended version of the paper titled \emph{Programmable Semantic Fragments}, which appears in the proceedings of the 15th International Conference on Generative Programming: Concepts \& Experience (GPCE 2016).}
+\abstract{This technical report serves to archive the extended version of the paper \emph{Programmable Semantic Fragments: The Design and Implementation of \texttt{typy}}, which appears in the proceedings of the 15th International Conference on Generative Programming: Concepts \& Experience (GPCE'16).}
 \begin{document}
